---
title: "Ma_2020_sarilumab_anc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ma_2020_sarilumab_anc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup}
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(PKNCA)
```

# Sarilumab PopPK/PD for absolute neutrophil count (ANC)

The model is a semi-mechanistic indirect-response PopPK/PD model in which
sarilumab concentrations **stimulate** the first-order elimination rate of
absolute neutrophil count (ANC) — mechanistically, margination of
functional neutrophils from circulation to the vascular wall or tissue
(Ma 2020). The sarilumab PK backbone is the two-compartment model with
first-order absorption and parallel linear + Michaelis–Menten elimination
published by Xu et al. 2019 (companion paper from the same sponsor).

* Ma 2020: [Clin Pharmacokinet 59(11):1451–1466](https://doi.org/10.1007/s40262-020-00899-7) (PMID 32451909).
* Xu 2019: [Clin Pharmacokinet 58(11):1455–1467](https://doi.org/10.1007/s40262-019-00765-1) (PMID 31055792).

```{r citation-and-population}
mod <- readModelDb("Ma_2020_sarilumab_anc")
cat(rxode2::rxode(mod)$reference, sep = "\n")
```

# Population

The ANC model was developed from 1672 patients with rheumatoid arthritis
pooled across one phase I (NCT01011959), one phase II (NCT01061736 Part
A), and three phase III studies (NCT01061736 Part B [MOBILITY],
NCT01709578 [TARGET], NCT01768572 [ASCERTAIN]); Ma 2020 Table 1. Baseline
demographics (Ma 2020 Table 2): mean age 51.7 years (SD 12.1); mean
weight 74.1 kg (SD 18.7); 82.2% female; 84.8% Caucasian; 14.2% smokers;
97.7% on concomitant methotrexate; 63.8% with prior corticosteroid
treatment; mean baseline ANC 5.38 × 10⁹/L (Table 4).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Ma_2020_sarilumab_anc.R`.
The table below collects them in one place for review.

| Equation / parameter | Value | Source location |
| --- | --- | --- |
| PK structure (2-cmt SC, linear + MM) | n/a | Xu 2019 Fig. 1 (base model scheme) |
| Vm (mg/day) | 8.06 | Xu 2019 Table 3 |
| Km (mg/L) | 0.939 | Xu 2019 Table 3 |
| Vc/F (L) | 2.08 | Xu 2019 Table 3 |
| CLO/F (L/day) | 0.260 | Xu 2019 Table 3 |
| Ka (1/day) | 0.136 | Xu 2019 Table 3 |
| Q/F (L/day) | 0.156 | Xu 2019 Table 3 |
| Vp/F (L) | 5.23 | Xu 2019 Table 3 |
| PK IIV (CV% on Vm, CLO, Vc, Ka; block Vm/CLO r=-0.566) | 32.4 / 55.3 / 37.3 / 32.1 | Xu 2019 Table 3 |
| PK residual error (log-scale variance → proportional) | σ² = 0.395 → propSd ≈ 0.629 | Xu 2019 Table 3 |
| PD structure (indirect response, stim on Kout) | Eff = Emax·C^γ/(EC50^γ+C^γ); dANC/dt = Kin − Kout·(1+Eff)·ANC | Ma 2020 Fig. 1 + methods text |
| Baseline ANC BASE (10⁹/L) | 5.38 | Ma 2020 Table 4 |
| Emax (unitless) | 1.50 | Ma 2020 Table 4 |
| EC50 (mg/L) | 10.3 | Ma 2020 Table 4 |
| **Kout (1/day)** | **2.11 (corrected from published typo "211"; see Assumptions and deviations)** | **Ma 2020 Table 4 (bootstrap median; published estimate 2.17)** |
| γ Hill coefficient | 0.862 | Ma 2020 Table 4 |
| Smoking on BASE (power) | 1.15 | Ma 2020 Table 4 |
| Weight on Kout (power, ref 71 kg) | 0.875 | Ma 2020 Table 4 (footnote b) |
| PRICORT on Emax (power) | 0.819 | Ma 2020 Table 4 |
| PD IIV (CV% on BASE, Emax, EC50, Kout, γ) | 32.1 / 61.9 / 36.9 / 227 / 80.4 | Ma 2020 Table 4 |
| PD residual error (proportional, CV%) | 28.2 | Ma 2020 Table 4 |

## Covariate column naming

| Source column | Canonical column used here |
| --- | --- |
| Body weight | `WT` (kg; reference 71 kg) |
| Smoking | `SMOKE` (new canonical; see `covariate-columns.md`) |
| PRICORT | `PRICORT` (new canonical; see `covariate-columns.md`) |

# Virtual cohort

The original observed data are not public. The cohort below draws
subjects whose covariate distributions approximate the Ma 2020 ANC
dataset (Table 2).

```{r virtual-pop}
set.seed(2020)
n_subj <- 100
regimens <- c("150 q2w", "200 q2w")

pop <- tibble(
  id      = seq_len(n_subj),
  WT      = pmax(40, pmin(140, round(rnorm(n_subj, mean = 74.1, sd = 18.7), 1))),
  SMOKE   = rbinom(n_subj, 1, 0.142),
  PRICORT = rbinom(n_subj, 1, 0.638),
  regimen = rep(regimens, length.out = n_subj)
) |>
  mutate(
    dose_mg = ifelse(regimen == "150 q2w", 150, 200)
  )
head(pop)
```

# Simulation

Build one event table per subject (dose every 14 days for 24 weeks, with
Cc and ANC sampled weekly) and solve each subject independently, then
concatenate. This per-subject loop sidesteps a pre-existing rxode2 issue
in which multi-endpoint models with more than two subjects in a single
`rxSolve` call abort in the C solver. Each call samples a fresh set of
IIV etas from the model's omega specification, so the pooled result is
equivalent to the typical "sample N subjects" workflow.

```{r simulate}
mod <- readModelDb("Ma_2020_sarilumab_anc")

sim_one <- function(sub, seed_offset = 0L) {
  ev <- rxode2::et(amt = sub$dose_mg, ii = 14, until = 168, cmt = "depot") |>
    rxode2::et(seq(0, 168, by = 7), cmt = "Cc") |>
    rxode2::et(seq(0, 168, by = 7), cmt = "ANC")
  ev_df <- as.data.frame(ev)
  ev_df$id      <- sub$id
  ev_df$WT      <- sub$WT
  ev_df$SMOKE   <- sub$SMOKE
  ev_df$PRICORT <- sub$PRICORT
  set.seed(2020L + sub$id + seed_offset)
  out <- rxode2::rxSolve(mod, ev_df, returnType = "data.frame")
  out$id <- sub$id
  out
}

sim <- pop |>
  dplyr::group_split(id) |>
  lapply(sim_one) |>
  dplyr::bind_rows() |>
  dplyr::left_join(dplyr::select(pop, id, regimen, dose_mg), by = "id")

# Build a single-id event table for later PKNCA dose-mapping.
events <- pop |>
  dplyr::group_split(id) |>
  lapply(function(sub) {
    ev <- rxode2::et(amt = sub$dose_mg, ii = 14, until = 168, cmt = "depot") |>
      rxode2::et(seq(0, 168, by = 7), cmt = "Cc") |>
      rxode2::et(seq(0, 168, by = 7), cmt = "ANC")
    ev_df <- as.data.frame(ev)
    ev_df$id      <- sub$id
    ev_df$WT      <- sub$WT
    ev_df$SMOKE   <- sub$SMOKE
    ev_df$PRICORT <- sub$PRICORT
    ev_df
  }) |>
  dplyr::bind_rows() |>
  dplyr::left_join(dplyr::select(pop, id, regimen), by = "id")

dim(sim)
head(sim)
```

For deterministic typical-value traces (no between-subject variability),
solve with `rxode2::zeroRe()`. Two typical subjects (one per regimen) fit
inside the working subset of the multi-endpoint solver.

```{r simulate-typical}
mod_typ <- rxode2::zeroRe(mod)

typ_template <- pop |>
  dplyr::distinct(regimen, dose_mg) |>
  dplyr::mutate(id = dplyr::row_number())

typ_events <- typ_template |>
  dplyr::group_split(id) |>
  lapply(function(sub) {
    ev <- rxode2::et(amt = sub$dose_mg, ii = 14, until = 168, cmt = "depot") |>
      rxode2::et(seq(0, 168, by = 1), cmt = "Cc") |>
      rxode2::et(seq(0, 168, by = 1), cmt = "ANC")
    ev_df <- as.data.frame(ev)
    ev_df$id <- sub$id
    ev_df$WT <- 71
    ev_df$SMOKE <- 0L
    ev_df$PRICORT <- 0L
    ev_df
  }) |>
  dplyr::bind_rows()

sim_typ <- rxode2::rxSolve(mod_typ, typ_events, returnType = "data.frame") |>
  dplyr::left_join(typ_template, by = "id")
```

# Replicate published figures

## Figure 5 — ANC time profiles (200 mg vs 150 mg q2w)

```{r figure-5-anc}
# Replicates Figure 5 of Ma 2020 (ANC time profiles, observed vs predicted,
# 200 mg q2w and 150 mg q2w). Black line: typical-value. Ribbon: simulated
# 5th/95th percentiles across the virtual cohort.
summary_anc <- sim |>
  dplyr::filter(!is.na(effect)) |>
  dplyr::group_by(time, regimen) |>
  dplyr::summarise(
    P5  = quantile(effect, 0.05),
    P50 = quantile(effect, 0.50),
    P95 = quantile(effect, 0.95),
    .groups = "drop"
  )

typ_anc <- sim_typ |> dplyr::filter(!is.na(effect))

ggplot() +
  geom_ribbon(data = summary_anc, aes(time, ymin = P5, ymax = P95, fill = regimen), alpha = 0.2) +
  geom_line(data = typ_anc, aes(time, effect, color = regimen), linewidth = 0.8) +
  labs(
    x = "Time (days)",
    y = expression(ANC~(10^9/L)),
    title = "Figure 5 — Simulated ANC vs. time by regimen",
    caption = "Replicates the structure of Figure 5 of Ma 2020 (ANC time profiles, 150 vs 200 mg q2w)."
  ) +
  theme_minimal()
```

## Sarilumab concentration — typical-value companion

```{r figure-cc}
typ_pk <- sim_typ |> dplyr::filter(!is.na(Cc))
ggplot(typ_pk, aes(time, Cc, color = regimen)) +
  geom_line(linewidth = 0.8) +
  labs(
    x = "Time (days)",
    y = "Sarilumab Cc (mg/L)",
    title = "Typical-value sarilumab PK (Xu 2019 backbone)",
    caption = "Companion PK driver for the Ma 2020 ANC PopPK/PD model."
  ) +
  theme_minimal()
```

# Endogenous / turnover validation

This is an indirect-response model with a circulating endogenous species
(neutrophils) rather than a simple drug-PK model, so the primary
validation targets are (1) steady-state baseline, (2) dose-response at
steady state, and (3) return to baseline after drug withdrawal. Classical
NCA on ANC would not be meaningful. The sarilumab PK leg is validated
with PKNCA below.

## Steady-state baseline (no drug)

With no sarilumab dose, ANC must sit at `BASE` for all time. Confirm:

```{r baseline-check}
ev_placebo <- rxode2::et(seq(0, 56, by = 1), cmt = "ANC")
ev_placebo$WT <- 71; ev_placebo$SMOKE <- 0; ev_placebo$PRICORT <- 0
baseline_sim <- rxode2::rxSolve(mod_typ, ev_placebo, returnType = "data.frame")
range(baseline_sim$effect, na.rm = TRUE)
```

## Dose-response at steady state

Published value: at median trough of 200 mg q2w the predicted ANC
reduction from baseline is **39%** (i.e. nadir ANC ≈ 0.61 × 5.38 ≈ 3.28
× 10⁹/L). For 150 mg q2w: **31%** (nadir ≈ 3.71).

```{r steady-state-check}
ss_ranges <- typ_anc |>
  dplyr::filter(time >= 84, time <= 168) |>  # steady state window: weeks 12-24
  dplyr::group_by(regimen) |>
  dplyr::summarise(
    nadir_anc    = min(effect),
    trough_anc   = dplyr::last(effect),
    mean_anc     = mean(effect),
    .groups      = "drop"
  ) |>
  dplyr::mutate(
    pct_reduction_from_base = round(100 * (5.38 - nadir_anc) / 5.38, 1),
    paper_reported_pct      = ifelse(regimen == "200 q2w", 39, 31)
  )
knitr::kable(ss_ranges, caption = "Simulated ANC nadir and percent reduction vs. Ma 2020 reported predictions.")
```

## Recovery after drug withdrawal

Ma 2020 reports that ANC returns to baseline approximately 2 weeks after a
single 200 mg subcutaneous dose. Confirm:

```{r recovery-check}
ev_single <- rxode2::et(amt = 200, cmt = "depot") |>
  rxode2::et(seq(0, 56, by = 1), cmt = "ANC")
ev_single$WT <- 71; ev_single$SMOKE <- 0; ev_single$PRICORT <- 0
recovery_sim <- rxode2::rxSolve(mod_typ, ev_single, returnType = "data.frame")

# Time at which ANC is back within 2% of baseline
within_baseline <- recovery_sim |>
  dplyr::filter(!is.na(effect), time > 5) |>
  dplyr::filter(abs(effect - 5.38) / 5.38 <= 0.02) |>
  dplyr::slice(1)
within_baseline
```

# PKNCA validation (sarilumab PK)

Run PKNCA on the sarilumab concentration leg. The paper reports that a
**two-fold** increase in steady-state AUC₀–₁₄ is observed when escalating
from 150 to 200 mg q2w (abstract of Xu 2019), consistent with saturation
of the non-linear Michaelis–Menten pathway.

```{r pknca-cc}
# rxSolve emits one row per observation record, so with both Cc and ANC
# endpoints requested at the same time point we get duplicated (id, time)
# rows with identical Cc values. Collapse before handing to PKNCA.
sim_pk <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, regimen) |>
  dplyr::distinct(id, time, regimen, .keep_all = TRUE)

conc_obj <- PKNCA::PKNCAconc(sim_pk, Cc ~ time | regimen + id)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, regimen) |>
  dplyr::distinct(id, time, regimen, .keep_all = TRUE)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | regimen + id)

# Steady-state interval: last dose (day 168-14 = 154) to 168
intervals <- data.frame(
  start = 154,
  end   = 168,
  cmax  = TRUE,
  cmin  = TRUE,
  auclast = TRUE,
  tmax  = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- PKNCA::pk.nca(nca_data)
nca_summary <- summary(nca_res)
knitr::kable(nca_summary,
             caption = "Steady-state NCA (day 154-168) for sarilumab by regimen.")
```

# Assumptions and deviations

* **Kout typo correction (Table 4).** Ma 2020 Table 4 prints the ANC
  turnover rate constant as
  > K_out, day⁻¹: Estimate 2.17 (%RSE 35.3, 95% CI 0.638–3.71);
  > Bootstrap median (95% CI) **211 (1.67–2.88)**.
  The bootstrap median "211" is almost certainly a decimal-point error:
  the accompanying 95% confidence interval (1.67–2.88) brackets 2.11 but
  not 211 (a 100× discrepancy). A neutrophil turnover rate constant of
  ~2/day (half-life ~8 h) is physiologically consistent with the
  literature on neutrophil margination/demargination; 211/day (half-life
  ~5 min) is not.
  This package implements `Kout = 2.11` (the corrected bootstrap median)
  so that the point estimate, the bootstrap CI, and the physiology all
  agree. The reported population-mean estimate (2.17, %RSE 35.3) is very
  close to 2.11 and produces indistinguishable simulated ANC profiles at
  the labelled doses. The IIV %RSE on Kout, also printed in Table 4 as
  "227 / bootstrap 232 (198–279)", is retained as published.
  This correction is documented inline above the `lkout` entry in
  `inst/modeldb/specificDrugs/Ma_2020_sarilumab_anc.R`. Readers who need
  to reproduce the (implausible) published bootstrap median exactly can
  replace `log(2.11)` with `log(211)` in that file, though the resulting
  ANC trajectory will not match Ma 2020 Figure 5.
* **PK backbone from Xu 2019, typical-value only.** The sarilumab PK
  leg uses Xu 2019's two-compartment + linear + Michaelis–Menten model
  at its reference (71 kg female, albumin 38 g/L, CrCl 100 mL/min, BLCRP
  14.2 mg/L, ADA-negative, DP1/DP3). Xu 2019 covariate effects on PK
  parameters (body weight, albumin, CrCl, BLCRP, ADA, DP2, sex) are
  **not** reproduced in this file; they live in the companion
  `Xu_2019_sarilumab.R` model. The sibling DAS28-CRP PD model
  (`Ma_2020_sarilumab_das28crp.R`) makes the same choice for consistency.
* **PK IIV retained (Xu 2019), PD IIV retained (Ma 2020).** Both sets
  of between-subject variability are carried in `ini()` so stochastic
  VPCs from this file are reasonable. Use `rxode2::zeroRe(mod)` to
  suppress IIV for typical-value plots.
* **Covariate distributions in the virtual cohort** are drawn
  independently from the marginal distributions in Ma 2020 Table 2
  (ANC dataset): 14.2% smokers, 63.8% with prior corticosteroid,
  weight ~ N(74.1, 18.7²) truncated to [40, 140]. Correlations between
  covariates are not modelled.
* **Race is not a covariate of this PD model** (84.8% Caucasian in
  Ma 2020 Table 2; no RACE_* column is required).
* **The companion DAS28-CRP model** (`Ma_2020_sarilumab_das28crp.R`,
  extracted separately) uses the DAS28-CRP column of Ma 2020 Table 3.
  Table 3 has **no** known decimal-point errors; the typo discussion
  above applies only to this ANC model and Table 4.
