---
title: "Chloroquine PK/PD in Plasmodium vivax (Abd-Rahman 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chloroquine PK/PD in Plasmodium vivax (Abd-Rahman 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Abd-Rahman 2020 characterised the population PK and PD of chloroquine and its
active metabolite desethylchloroquine in a *Plasmodium vivax* induced
blood-stage malaria volunteer infection study. Chloroquine and
desethylchloroquine PK was described by a joint two-compartment
parent-metabolite model with first-order absorption and elimination, fit
**separately** to plasma and to whole-blood drug concentrations. Parasite
killing was described by a delayed-effect (effect-compartment) model in which
chloroquine in a hypothetical biophase drives a sigmoid Emax kill of the
log10 *P. vivax* parasitaemia.

nlmixr2lib ships the two matrices as two model files, both used below:

* Plasma model: `AbdRahman_2020_chloroquine_plasma`
* Whole-blood model: `AbdRahman_2020_chloroquine_wholeblood`

* Citation: `r rxode2::rxode(readModelDb("AbdRahman_2020_chloroquine_plasma"))$reference`
* Article: <https://doi.org/10.1002/cpt.1893> (open access)
* Supplement (methods, equations S1): <https://doi.org/10.1002/cpt.1893>

All concentrations and doses are in **molar** units, following the paper, which
converted chloroquine and desethylchloroquine to micromoles using molecular
masses of 319.8 and 291.8 g/mol. A chloroquine base dose in mg is converted to
micromoles as `mg / 319.8 * 1000`.

# Population

The analysis pooled 24 healthy malaria-naive adults (13 male, 11 female; mean
age 25.6 years, range 19-44; mean weight 73.3 kg, range 57.2-99.5) inoculated
intravenously with approximately 564 viable blood-stage *P. vivax*-infected
erythrocytes on day 0 (Abd-Rahman 2020 Table 1). Chloroquine phosphate
(Avloclor) was given as a total base dose of 1.55 g over three days for adults
>= 60 kg (25 mg base/kg for < 60 kg): 620 mg base at 0 h, then 310 mg base at 6,
24 and 48 h. Age, sex and body weight were screened as covariates but none was
retained in the final model.

The population metadata are available programmatically via
`readModelDb("AbdRahman_2020_chloroquine_plasma")()$population`.

# Source trace

Per-parameter provenance is recorded in-file next to every `ini()` entry in
`inst/modeldb/specificDrugs/AbdRahman_2020_chloroquine_plasma.R` and
`..._wholeblood.R`. The table collects the structural values (plasma /
whole-blood).

| Equation / parameter | Value (plasma / whole blood) | Source |
|---|---|---|
| `lka` (ka) | 0.943 / 0.574 1/h | Table 2 |
| `lcl` (CL_CQ/F) | 54.6 / 8.96 L/h | Table 2 |
| `lvc` (Vc_CQ/F) | 2930 / 560 L | Table 2 |
| `lq` (Q1_CQ/F) | 47.2 / 38.5 L/h | Table 2 |
| `lvp` (Vp1_CQ/F) | 4700 / 1230 L | Table 2 |
| `lfdepot` (Frel) | 1 (fixed) | Table 2 |
| `lcl_dcq` (CL_DCQ/F) | 37.6 / 4.42 L/h | Table 2 |
| `lvc_dcq` (Vc_DCQ/F) | 40.0 / 16.1 L | Table 2 |
| `lq_dcq` (Q1_DCQ/F) | 36.3 / 4.46 L/h | Table 2 |
| `lvp_dcq` (Vp1_DCQ/F) | 2840 / 259 L | Table 2 |
| `fm` | 0.18 (fixed) | Results / refs 9,12 |
| `lemax` (EmaxCQ) | 0.213 1/h (shared) | Table 3 |
| `lec50` (EC50) | 0.047 / 0.28 umol/L (fixed) | Table 3 |
| `lhill` (Hill) | 2.5 (fixed) | Table 3 |
| `lke0` (ke0) | 0.0212 / 0.0288 1/h | Table 3 |
| `lkgrow` (kgrow) | 0.059 1/h (shared) | Table 3 |
| `plbase` (PLbase) | -3.36 log10/mL (shared) | Table 3 |
| `d/dt(central)` etc. (2-cmt CQ + DCQ) | n/a | Figure 1 |
| `dCe/dt = ke0*(Cc-Ce)`; `kkill = Emax*Ce^Hill/(EC50^Hill+Ce^Hill)` | n/a | Suppl. S1 |
| `dPL/dt = kgrow - kkill` (log parasitaemia) | n/a | Suppl. S1 |

The parasite ODE deserves a note. Supplementary Material S1 writes
`dPL/dt = kgrow - kkill` with `PL` the log-transformed parasite count, but the
same supplement's derived quantities use natural *e*
(`PMR48 = exp(kgrow*48) = 17.4` and `PCt1/2 = ln(2)/(Emax-kgrow) = 4.5 h`), so
`kgrow` and `Emax` are natural-log growth/kill rates. The packaged models store
the state in log10 (so the reported baseline `PLbase = -3.36` and the additive
residual on log10 parasitaemia apply directly), which makes the ODE
`dPL10/dt = (kgrow - kkill)/ln(10)`. This is algebraically identical to a
natural-log implementation and reproduces every reported secondary parameter
(checked below).

# Virtual cohort

Original observed data are not public. The cohort below approximates the trial:
100 subjects per matrix given the paper's 3-day regimen. `set.seed()` fixes R's
RNG but not rxode2's per-thread simulation stream, so the exact draw differs
across machines; every assertion below is written to hold for any cohort the
model can produce (centre and robust quantiles, not extremes).

```{r cohort}
set.seed(74)

mw_cq <- 319.8
mg_to_umol <- function(mg) mg / mw_cq * 1000
dose_umol <- mg_to_umol(c(620, 310, 310, 310))
dose_times <- c(0, 6, 24, 48)

n_sub <- 100L
# Dense sampling early (absorption / distribution) and coarse late (the long
# terminal phase; chloroquine t1/2 ~ 150 h needs a long tail for NCA).
obs_times <- sort(unique(c(seq(0, 72, by = 0.5), seq(78, 1200, by = 6))))

make_events <- function(n, id_offset = 0L) {
  ids <- id_offset + seq_len(n)
  doses <- tidyr::crossing(id = ids, dn = seq_along(dose_times)) |>
    mutate(
      time = dose_times[dn], amt = dose_umol[dn],
      evid = 1L, cmt = "depot", dvid = NA_integer_
    ) |>
    select(-dn)
  obs <- tidyr::crossing(id = ids, time = obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L)
  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}

events <- make_events(n_sub)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod_plasma <- readModelDb("AbdRahman_2020_chloroquine_plasma")
mod_wb <- readModelDb("AbdRahman_2020_chloroquine_wholeblood")

# useLinCmt = FALSE: rxode2's automatic ODE -> linCmt conversion corrupts the
# dvid -> cmt mapping for multi-output models such as these.
sim_plasma <- rxode2::rxSolve(mod_plasma, events = events, useLinCmt = FALSE) |>
  as.data.frame() |>
  mutate(matrix = "plasma")
sim_wb <- rxode2::rxSolve(mod_wb, events = events, useLinCmt = FALSE) |>
  as.data.frame() |>
  mutate(matrix = "whole blood")
sim <- bind_rows(sim_plasma, sim_wb)
```

Typical-value profiles (no between-subject variability) for the figure
replications and the closed-form PD checks. `zeroRe()` segfaults on
multi-endpoint models, so the random effects are zeroed by setting the eta
columns to 0 and passing `omega = NA`.

```{r simulate-typical}
eta_cols_plasma <- c(
  "etalka", "etalfdepot", "etalcl", "etalvc", "etalcl_dcq",
  "etaplbase", "etalkgrow", "etalemax", "etalec50", "etalke0"
)
eta_cols_wb <- c(eta_cols_plasma, "etalvc_dcq")

events_typ <- make_events(1L)

typ_plasma <- rxode2::rxSolve(
  mod_plasma,
  events = dplyr::bind_cols(
    events_typ,
    as.data.frame(setNames(as.list(rep(0, length(eta_cols_plasma))), eta_cols_plasma))[rep(1, nrow(events_typ)), ]
  ),
  omega = NA, useLinCmt = FALSE
) |> as.data.frame()

typ_wb <- rxode2::rxSolve(
  mod_wb,
  events = dplyr::bind_cols(
    events_typ,
    as.data.frame(setNames(as.list(rep(0, length(eta_cols_wb))), eta_cols_wb))[rep(1, nrow(events_typ)), ]
  ),
  omega = NA, useLinCmt = FALSE
) |> as.data.frame()
```

# Replicate published figures

## Figure 2 -- PK visual predictive check

Figure 2 of Abd-Rahman 2020 shows VPCs of chloroquine and desethylchloroquine
in plasma and whole blood. The panels below reproduce the 5th/50th/95th
simulated percentiles over the first 480 h.

```{r figure-2}
sim |>
  filter(time <= 480) |>
  select(id, time, matrix, Chloroquine = Cc, Desethylchloroquine = Cc_dcq) |>
  pivot_longer(c(Chloroquine, Desethylchloroquine),
    names_to = "analyte", values_to = "conc"
  ) |>
  group_by(matrix, analyte, time) |>
  summarise(
    Q05 = quantile(conc, 0.05, na.rm = TRUE),
    Q50 = quantile(conc, 0.50, na.rm = TRUE),
    Q95 = quantile(conc, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(Q50 > 0) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_grid(analyte ~ matrix, scales = "free_y") +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Concentration (umol/L)",
    title = "Figure 2 -- chloroquine / desethylchloroquine VPC",
    caption = "Replicates Figure 2 of Abd-Rahman 2020."
  )
```

## Figure 3 -- parasitaemia clearance

Figure 3 shows the *P. vivax* parasitaemia decline after chloroquine treatment.
The typical-value trajectories below start at the fitted baseline
(`PLbase = -3.36` log10 parasites/mL), decline while chloroquine holds the
effect site above EC50, and begin to regrow once the effect-site concentration
falls below EC50.

```{r figure-3}
bind_rows(
  typ_plasma |> transmute(time, matrix = "plasma", pl = parasitemia_log10),
  typ_wb |> transmute(time, matrix = "whole blood", pl = parasitemia_log10)
) |>
  filter(time <= 700) |>
  ggplot(aes(time, pl, colour = matrix)) +
  geom_line() +
  labs(
    x = "Time after first dose (h)", y = "log10 parasitaemia (parasites/mL)",
    title = "Figure 3 -- P. vivax parasitaemia clearance (typical value)",
    caption = "Replicates Figure 3 of Abd-Rahman 2020."
  )
```

# PKNCA validation

NCA of the simulated cohort against the secondary PK parameters in Abd-Rahman
2020 Table 2 (reported as median across subjects). We compute per-matrix,
per-analyte Cmax, Tmax, AUC0-inf and terminal half-life and compare the cohort
**medians** to the published medians.

```{r pknca}
run_nca <- function(simdf, conc_col) {
  nca_in <- simdf |>
    rename(conc = all_of(conc_col)) |>
    filter(!is.na(conc)) |>
    select(id, time, conc)
  # Guarantee a time=0 row per subject (extravascular pre-dose conc = 0).
  nca_in <- bind_rows(
    nca_in,
    nca_in |> distinct(id) |> mutate(time = 0, conc = 0)
  ) |>
    distinct(id, time, .keep_all = TRUE) |>
    arrange(id, time)

  conc_obj <- PKNCA::PKNCAconc(nca_in, conc ~ time | id)
  dose_df <- events |>
    filter(evid == 1L) |>
    select(id, time, amt)
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | id)
  intervals <- data.frame(
    start = 0, end = Inf,
    cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
  )
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
  as.data.frame(res$result)
}

nca_all <- bind_rows(
  run_nca(sim_plasma, "Cc") |> mutate(matrix = "plasma", analyte = "chloroquine"),
  run_nca(sim_plasma, "Cc_dcq") |> mutate(matrix = "plasma", analyte = "desethylchloroquine"),
  run_nca(sim_wb, "Cc") |> mutate(matrix = "whole blood", analyte = "chloroquine"),
  run_nca(sim_wb, "Cc_dcq") |> mutate(matrix = "whole blood", analyte = "desethylchloroquine")
)

nca_med <- nca_all |>
  group_by(matrix, analyte, PPTESTCD) |>
  summarise(simulated = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life"))
```

## Comparison against published NCA

```{r nca_compare}
published <- tibble::tribble(
  ~matrix, ~analyte, ~PPTESTCD, ~reference,
  "plasma", "chloroquine", "cmax", 0.72,
  "plasma", "chloroquine", "tmax", 4.0,
  "plasma", "chloroquine", "aucinf.obs", 80,
  "plasma", "chloroquine", "half.life", 149,
  "plasma", "desethylchloroquine", "cmax", 0.08,
  "plasma", "desethylchloroquine", "tmax", 3.3,
  "plasma", "desethylchloroquine", "aucinf.obs", 21,
  "plasma", "desethylchloroquine", "half.life", 104,
  "whole blood", "chloroquine", "cmax", 2.83,
  "whole blood", "chloroquine", "tmax", 4.0,
  "whole blood", "chloroquine", "aucinf.obs", 509,
  "whole blood", "chloroquine", "half.life", 156,
  "whole blood", "desethylchloroquine", "cmax", 0.39,
  "whole blood", "desethylchloroquine", "tmax", 4.0,
  "whole blood", "desethylchloroquine", "aucinf.obs", 233,
  "whole blood", "desethylchloroquine", "half.life", 83
)

cmp <- nca_med |>
  left_join(published, by = c("matrix", "analyte", "PPTESTCD")) |>
  mutate(
    parameter = recode(PPTESTCD,
      cmax = "Cmax (umol/L)", tmax = "Tmax (h)",
      aucinf.obs = "AUC0-inf (umol*h/L)", half.life = "t1/2 (h)"
    ),
    pct_diff = 100 * (simulated - reference) / reference,
    flag = ifelse(abs(pct_diff) > 25, "*", "")
  ) |>
  arrange(matrix, analyte, parameter)

cmp |>
  transmute(
    Matrix = matrix, Analyte = analyte, Parameter = parameter,
    Simulated = round(simulated, 3), Published = reference,
    `% diff` = round(pct_diff, 1), Flag = flag
  ) |>
  knitr::kable(
    caption = "Simulated (cohort median) vs. published median NCA. * differs by >25%."
  )
```

The chloroquine PK is well reproduced: terminal half-life agrees to within 2% in
both matrices (the strongest discriminator of the two-compartment disposition),
and AUC0-inf and Cmax agree to within about 12%. The starred rows are known,
explained deviations and were not tuned away:

* **Chloroquine Tmax** is about 9 h against the published 4 h. Under the
  four-dose regimen the simulated global peak falls just after the 6 h dose,
  where the second dose adds to the first. The published Tmax probably
  describes the first-dose absorption peak. The paper does not define how Tmax
  was derived from the empirical-Bayes profiles. The Cmax *magnitude* agrees.
* **Desethylchloroquine half-life.** In this model desethylchloroquine
  elimination (`kel_dcq` of about 0.9 1/h in plasma) is much faster than its
  formation, so the metabolite is formation-rate limited. Its *observed*
  terminal half-life therefore has to follow chloroquine's (about 150 h), which
  is what the NCA returns. The published 104 h (plasma) and 83 h (whole blood)
  are shorter than the parent's, so they cannot be observed terminal values.
  They match the metabolite's own two-compartment disposition half-life computed
  from its Table 2 parameters (checked in the next chunk). That agreement also
  confirms that the metabolite parameters were transcribed correctly.
* **Desethylchloroquine Tmax and Cmax.** The simulated metabolite accumulates
  over the dosing days and peaks near the last dose (about 50 h). The paper
  reports a metabolite Tmax of 3-4 h, the same as the parent's, and a lower Cmax,
  and it does not describe the calculation. Metabolite exposure (AUC0-inf) agrees
  within 20%, and that is the quantity the fixed FM = 0.18 governs.

```{r nca_gate}
# Gate on the structural discriminators that the model is expected to reproduce
# tightly: chloroquine half-life and AUC in both matrices. Assert on the centre
# (cohort median vs published median), not on extremes.
chk <- cmp |> filter(analyte == "chloroquine", PPTESTCD %in% c("half.life", "aucinf.obs"))
stopifnot(nrow(chk) == 4)
stopifnot(all(abs(chk$pct_diff) < 25))
```

```{r dcq_intrinsic_half_life}
# Desethylchloroquine intrinsic two-compartment disposition half-life from its
# own Table 2 parameters: the slow eigenvalue of the metabolite sub-system.
# Deterministic (no simulation), so a tight bound is appropriate.
beta_half_life <- function(mod) {
  ini <- mod()$iniDf
  g <- function(nm) exp(ini$est[ini$name == nm])
  k10 <- g("lcl_dcq") / g("lvc_dcq")
  k12 <- g("lq_dcq") / g("lvc_dcq")
  k21 <- g("lq_dcq") / g("lvp_dcq")
  a <- k10 + k12 + k21
  beta <- (a - sqrt(a^2 - 4 * k10 * k21)) / 2
  log(2) / beta
}
dcq_thalf <- c(plasma = beta_half_life(mod_plasma), wholeblood = beta_half_life(mod_wb))
knitr::kable(
  tibble::tibble(
    Matrix = c("plasma", "whole blood"),
    `Intrinsic DCQ t1/2 (h)` = round(dcq_thalf, 1),
    `Published DCQ t1/2 (h)` = c(104, 83)
  ),
  caption = "Desethylchloroquine disposition half-life from its own parameters vs Table 2."
)
stopifnot(abs(dcq_thalf[["plasma"]] - 104) / 104 < 0.10)
stopifnot(abs(dcq_thalf[["wholeblood"]] - 83) / 83 < 0.10)
```

# Pharmacodynamic validation

The PD secondary parameters have closed-form relationships to `kgrow` and
`Emax` that the packaged model reproduces exactly.

```{r pd-secondary}
p_plasma <- mod_plasma()
ini_p <- p_plasma$iniDf
getv <- function(df, nm) df$est[df$name == nm]
kgrow <- exp(getv(ini_p, "lkgrow"))
emax <- exp(getv(ini_p, "lemax"))

pmr48 <- exp(kgrow * 48) # parasite multiplication rate per 48 h
pct_half <- log(2) / (emax - kgrow) # parasite clearance half-life (h)
log10_prr48_max <- (emax - kgrow) / log(10) * 48 # max log10 PRR over 48 h

pd_tab <- tibble::tibble(
  Parameter = c("PMR48 (fold/48 h)", "PCt1/2 (h)", "max log10 PRR48"),
  Simulated = round(c(pmr48, pct_half, log10_prr48_max), 2),
  Published = c("17.4", "4.5", "~2.6 (obs), 3.2 (max)")
)
knitr::kable(pd_tab, caption = "PD secondary parameters vs Abd-Rahman 2020 Table 3.")

stopifnot(abs(pmr48 - 17.4) < 1.0) # paper 17.4 (95% CI 14.3-20.4)
stopifnot(abs(pct_half - 4.5) < 0.3) # paper 4.5 h (95% CI 4.1-5.0)
```

Baseline parasitaemia and the delayed-kill dynamics also reproduce the paper.
The typical-value trajectory starts at the fitted `PLbase = -3.36`, and the
time for which the chloroquine effect site stays above EC50 is close to the
paper's reported "time above EC50" (geometric means of 14.3 days in plasma and
18.3 days in whole blood). The typical subject runs about 3 days longer (17.8 and
21.2 days). The published values are geometric means across individual
empirical-Bayes profiles, and each subject has its own EC50 (fixed 30% IIV).

```{r pd-dynamics}
stopifnot(abs(typ_plasma$parasitemia_log10[1] - (-3.36)) < 1e-6)
stopifnot(abs(typ_wb$parasitemia_log10[1] - (-3.36)) < 1e-6)

ec50_plasma <- exp(getv(ini_p, "lec50"))
ec50_wb <- exp(getv(mod_wb()$iniDf, "lec50"))

time_above <- function(df, ec50) {
  d <- df[df$time > 0, ]
  above <- d$effect > ec50
  if (!any(above)) {
    return(0)
  }
  max(d$time[above]) / 24 # days (contiguous while chloroquine sustains the effect site)
}
ta_plasma <- time_above(typ_plasma, ec50_plasma)
ta_wb <- time_above(typ_wb, ec50_wb)

tibble::tibble(
  Matrix = c("plasma", "whole blood"),
  `Time above EC50 (days), simulated` = round(c(ta_plasma, ta_wb), 1),
  `Published (days)` = c(14.3, 18.3)
) |>
  knitr::kable(caption = "Time above EC50 vs Abd-Rahman 2020 (geometric mean).")

# Typical-value (deterministic) time above EC50 was 17.8 d (plasma) and
# 21.2 d (whole blood) when authored, against published geometric means of 14.3 d
# and 18.3 d. The window below catches a wrong ke0, EC50 or PK scale, each of
# which shifts this by many days.
stopifnot(abs(ta_plasma - 14.3) < 5)
stopifnot(abs(ta_wb - 18.3) < 5)
```

# Assumptions and deviations

- **Molar units.** All concentrations and doses are in micromoles / (umol/L),
  as in the paper. Convert a chloroquine base dose in mg with `mg / 319.8 * 1000`.
- **Log-scale parasite ODE.** Supplementary Material S1 writes
  `dPL/dt = kgrow - kkill` with `PL` a log-transformed parasite count, but its
  own secondary-parameter formulas use natural *e*. The maintainers store the
  state in log10 and use `dPL10/dt = (kgrow - kkill)/ln(10)`, which is
  algebraically identical to a natural-log implementation and reproduces every
  reported secondary parameter (PMR48, PCt1/2, PRR48). The reported baseline
  `PLbase = -3.36 log10 parasites/mL` is used directly as the state initial
  condition and its normal (not log-normal) between-subject distribution is
  encoded as an additive random effect.
- **Fixed FM = 0.18.** The fraction of chloroquine converted to
  desethylchloroquine was not identifiable and was fixed at 0.18 from urinary
  recovery literature (paper references 9 and 12). Molar 1:1 conversion is
  assumed (desethylchloroquine is chloroquine less an ethyl group).
- **Fixed EC50 and Hill.** No subject recrudesced, so the whole-blood EC50 was
  fixed to a literature relapse-based value (0.28 umol/L), the plasma EC50 to one
  sixth of it (0.047 umol/L, the whole-blood:plasma ratio), and the Hill
  coefficient to 2.5. Their between-subject variances were also fixed (Table 3).
- **Relative bioavailability.** Frel was fixed to 100% with estimated
  between-subject variability (Table 2).
- **Two model files.** The paper fit PK separately to plasma and whole-blood
  data and fit a single PD model with matrix-specific EC50 and ke0. The
  maintainers ship one full PK/PD model per matrix so each is self-contained;
  Emax, kgrow and PLbase are shared between the two files, EC50 and ke0 differ.
- **Covariates.** Age, sex and body weight were screened but not retained in the
  final model; they are recorded as documented-but-unused metadata.
- **NCA deviations.** Chloroquine Tmax and the desethylchloroquine Tmax, Cmax
  and observed half-life differ from the published medians for the reasons given
  in the NCA section (multi-dose accumulation, formation-rate-limited metabolite
  kinetics, and a published metabolite half-life that is its intrinsic disposition
  value). None was tuned.
- No correction or erratum for this article was found as of 2026-09-26.
