---
title: "Caspofungin (Abdul-Aziz 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Caspofungin (Abdul-Aziz 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("AbdulAziz_2025_caspofungin"))$reference`
* Description: `r rxode2::rxode(readModelDb("AbdulAziz_2025_caspofungin"))$description`
* Article: <https://doi.org/10.1128/aac.01435-24>
* Open-access full text: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC11823646/>

This is a one-compartment model with first-order elimination for intravenous
caspofungin in critically ill adults on extracorporeal membrane oxygenation
(ECMO), from the ASAP ECMO study program. It is a deliberately small model: two
structural parameters, one covariate, and an additive residual error. That
makes it unusually amenable to exact verification, and the sections below take
advantage of that -- every structural claim is checked against a closed form
rather than against a picture.

# Population

Eight critically ill adults contributed 64 plasma concentration-time points,
sampled over a **single dosing interval** once the patient had been stabilised
on ECMO (Abdul-Aziz 2025 Results, "Study population characteristics"). The
cohort median age was 36.5 years (range 20.0-62.0), actual body weight 84.0 kg
(range 55.0-130.0), and lean body weight 58.9 kg (range 37.3-81); 5 of 8
(62.5%) were male (Table 1). All patients were hypoalbuminaemic (albumin
17-33 g/L) and renal function spanned impairment to markedly augmented
clearance (Cockcroft-Gault creatinine clearance median 57.0 mL/min, range
31.0-260.0).

Every patient was cannulated onto ECMO -- veno-venous in 6 (75.0%) and
veno-arterial in 2 (25.0%) -- and 5 of 8 (62.5%) received concomitant renal
replacement therapy during sampling (4 CVVHD, 1 SLED). ECMO support is
therefore a property of this population rather than a covariate in the model:
the indicator is constant at 1 and cannot be estimated. The ECMO variables that
*do* vary within the cohort (duration, mode, flow rate) were all screened and
none was retained, which is the paper's headline finding -- "the use of ECMO had
negligible impact on the pharmacokinetics of caspofungin in critically ill
patients" (Discussion).

All patients received a 70 mg loading dose on day 1 as a 1 h infusion, followed
by a daily maintenance dose of 50 mg (6 patients) or 70 mg (2 patients). The
virtual cohort below reproduces that 6:2 split.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("AbdulAziz_2025_caspofungin")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/AbdulAziz_2025_caspofungin.R`. The table
below collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL) | 0.55 L/h | Table 2, "CL (L/h)" = 0.55 (%RSE 11.9); bootstrap median 0.55 (95% CI 0.42-0.68); restated in the Abstract |
| `lvc` (V at FFM = 58.9 kg) | 8.13 L | Table 2, "V (L)" = 8.13 (%RSE 7.15); bootstrap median 8.27 (95% CI 7.07-10.8); restated in the Abstract |
| `e_ffm_vc` | 1.24 | Table 2, "LBW effect on V" = 1.24 (%RSE 23.3); bootstrap median 1.24 (95% CI 0.08-1.91) |
| `etalcl` (BSV on CL) | 0.062996 = log(1 + 0.255^2) | Table 2, BSV "CL (%)" = 25.5 (%RSE 34.3) |
| `etalvc` (BSV on V) | 0.0054758 = log(1 + 0.0741^2) | Table 2, BSV "V (%)" = 7.41 (%RSE 101) |
| `addSd` | 1.20 mg/L | Table 2, Residual error "Additive (mg/L)" = 1.20 (%RSE 10.0) |
| `vc <- exp(lvc + etalvc) * (FFM / 58.9)^e_ffm_vc` | n/a | Displayed equation in Results, "Covariate screening and model development": `V = Vpop * (LBW / 58.9)^1.24`. The 58.9 kg divisor is the cohort median lean body weight (Table 1), per the Methods statement that continuous covariates were centred on their medians |
| `cl <- exp(lcl + etalcl)` | n/a | Methods, "Structural model development": BSV "described using an exponential model" `theta_j = theta_p * exp(eta_j)`. No covariate retained on clearance (serum albumin was tested and rejected) |
| `d/dt(central) <- -kel * central` | n/a | Results, "Population pharmacokinetic model building": "best described by a one-compartment model with first-order elimination"; a two-compartment model did not reduce BICc |
| `Cc ~ add(addSd)` | n/a | Results: "Residual unexplained variability was described by an additive error model" |
| Reference FFM = 58.9 kg | 58.9 kg | Table 1, "Lean body weight (in kg)" median 58.9 (range 37.3-81); also the divisor printed in the covariate equation |

The `LBW` label used throughout the paper is registered here as the canonical
covariate `FFM`. The register discriminates fat-free mass from lean body mass by
the estimating *formula*, not by the paper's label, and Abdul-Aziz 2025 cites
Janmahasatian et al. -- the fat-free-mass equation. The paper itself makes the
identification explicit: "Although the Janmahasatian et al. equation was
developed to estimate fat-free mass, the estimate is considered to be a
representation of lean body weight and is commonly used interchangeably"
(Discussion). This follows the identical `LBW` -> `FFM` precedent in
`Rolsma_2025_cefepime.R`.

# Structural verification against closed forms

A one-compartment model with first-order elimination has exact analytic
identities, so these checks compare the packaged model against arithmetic
derived from the paper's *printed* numbers (0.55 L/h, 8.13 L, exponent 1.24,
58.9 kg) rather than against a digitised figure. They are deterministic --
typical values with the random effects zeroed -- so the tolerances are tight on
purpose: the residual difference is pure numerical error, not cohort sampling.

```{r closed-forms}
mod <- readModelDb("AbdulAziz_2025_caspofungin")
mod_typ <- rxode2::zeroRe(mod)

# The paper's printed values, transcribed once and reused below.
pub_cl  <- 0.55  # L/h,  Table 2
pub_v   <- 8.13  # L,    Table 2 (at FFM = 58.9 kg)
pub_exp <- 1.24  # -,    Table 2 "LBW effect on V"
pub_ref <- 58.9  # kg,   Table 1 median lean body weight

# Build a typical-value event table: 70 mg loading dose then `md` mg q24h,
# each as a 1 h intravenous infusion into the central compartment.
make_events <- function(md, ffm, ndays, by = 0.02, id = 1L,
                        obs_end = 24 * ndays) {
  amts <- c(70, rep(md, ndays - 1L))
  dose <- data.frame(
    id = id, time = seq(0, by = 24, length.out = ndays),
    amt = amts, rate = amts, # rate = amt / 1 h infusion
    evid = 1L, cmt = "central", FFM = ffm
  )
  obs <- data.frame(
    id = id, time = seq(0, obs_end, by = by),
    amt = NA_real_, rate = NA_real_,
    evid = 0L, cmt = "central", FFM = ffm
  )
  rbind(dose, obs)
}
```

## Fat-free-mass power scaling

The volume must equal `8.13 * (FFM / 58.9)^1.24` at every fat-free mass, and
must equal exactly 8.13 L at the reference. Checked across the full observed
range of the cohort (37.3-81 kg) plus the four values the paper simulated
(40, 50, 60, 80 kg).

```{r ffm-scaling}
ffm_grid <- c(37.3, 40, 50, 58.9, 60, 80, 81)
ffm_chk <- vapply(ffm_grid, function(f) {
  s <- rxode2::rxSolve(mod_typ, make_events(50, f, ndays = 1L, by = 12),
                       returnType = "data.frame")
  unique(s$vc)[1]
}, numeric(1))

ffm_tab <- data.frame(
  FFM = ffm_grid,
  model = ffm_chk,
  closed_form = pub_v * (ffm_grid / pub_ref)^pub_exp
)
ffm_tab$rel_diff <- abs(ffm_tab$model - ffm_tab$closed_form) / ffm_tab$closed_form

ffm_tab |>
  dplyr::rename(
    "FFM (kg)"                    = FFM,
    "Model Vc (L)"                = model,
    "8.13 * (FFM/58.9)^1.24 (L)"  = closed_form,
    "Relative difference"         = rel_diff
  ) |>
  knitr::kable(digits = c(1, 4, 4, 12),
               caption = "Central volume vs the paper's displayed covariate equation.")

# Deterministic identity: tolerance is floating-point noise only.
stopifnot(max(ffm_tab$rel_diff) < 1e-8)
# The reference volume must be the printed typical value.
stopifnot(abs(ffm_tab$model[ffm_tab$FFM == pub_ref] - pub_v) < 1e-9)
```

## Terminal half-life

For a one-compartment model, `t1/2 = log(2) * V / CL`. Recovered here by
regressing `log(Cc)` on time over a window well clear of the last infusion, so
the estimate comes from the *solved concentrations* rather than from the model's
own `kel` output column (a gate built from a model variable cannot go red).

```{r half-life}
s_hl <- rxode2::rxSolve(
  mod_typ,
  # Dose for 5 days, then observe well past the last dose so a terminal
  # window is actually available to regress over.
  make_events(50, pub_ref, ndays = 5L, by = 0.25, obs_end = 220),
  returnType = "data.frame"
)
# Last infusion ends at 97 h; regress 130-200 h (>3 half-lives clear of it).
w_hl <- s_hl[!is.na(s_hl$Cc) & s_hl$time >= 130 & s_hl$time <= 200, ]
stopifnot(nrow(w_hl) > 50) # a regression on zero rows must not pass silently
kel_fit <- -unname(coef(lm(log(Cc) ~ time, data = w_hl))[2])

hl_model <- log(2) / kel_fit
hl_closed <- log(2) * pub_v / pub_cl
c(measured_h = hl_model, closed_form_h = hl_closed,
  rel_diff = abs(hl_model - hl_closed) / hl_closed)

stopifnot(abs(hl_model - hl_closed) / hl_closed < 1e-6)
```

## Mass-balance AUC identity

For any linear model with elimination only from the central compartment,

```
AUC(0, T) == (amount infused by T - amount remaining at T) / CL
```

holds **exactly at any T**, in or out of steady state. That makes it the
strongest available gate for a paper whose sampling covered only a single
dosing interval: it needs no steady-state assumption and no extrapolation. A
mis-transcribed clearance, dose, infusion duration or ODE sign breaks it
immediately.

```{r mass-balance}
s_mb <- rxode2::rxSolve(mod_typ, make_events(50, pub_ref, ndays = 6L, by = 0.02),
                        returnType = "data.frame")
o_mb <- s_mb[!is.na(s_mb$Cc), ]

trap <- function(t, y) sum(diff(t) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)

# Amount actually infused by time T, valid mid-infusion as well as after it.
infused_by <- function(T, dose_times, dose_amts, dur = 1) {
  sum(dose_amts * pmin(pmax(T - dose_times, 0), dur) / dur)
}
dose_times <- seq(0, by = 24, length.out = 6L)
dose_amts <- c(70, rep(50, 5L))

mb <- do.call(rbind, lapply(c(0.5, 6, 12, 24, 36, 48, 100, 144), function(T) {
  w <- o_mb[o_mb$time <= T, ]
  lhs <- trap(w$time, w$Cc)
  rhs <- (infused_by(T, dose_times, dose_amts) - w$central[nrow(w)]) / pub_cl
  data.frame(T = T, auc_trapezoid = lhs, auc_mass_balance = rhs,
             rel_diff = abs(lhs - rhs) / rhs)
}))

mb |>
  dplyr::rename(
    "T (h)"                      = T,
    "AUC(0,T) trapezoid (mg.h/L)" = auc_trapezoid,
    "(infused - remaining)/CL"    = auc_mass_balance,
    "Relative difference"         = rel_diff
  ) |>
  knitr::kable(digits = c(1, 4, 4, 10),
               caption = "Mass-balance AUC identity, exact at every T including mid-infusion (T = 0.5 h).")

# Deterministic identity; the residual is trapezoidal error on a 0.02 h grid.
stopifnot(max(mb$rel_diff) < 1e-5)
```

## Steady-state AUC over a dosing interval

At steady state `AUC(0, tau) == Dose / CL`, so the typical patient's day-to-day
exposure on 50 mg and 70 mg maintenance dosing is pinned by clearance alone.

```{r steady-state}
ss_chk <- do.call(rbind, lapply(c(50, 70), function(md) {
  s <- rxode2::rxSolve(mod_typ, make_events(md, pub_ref, ndays = 12L, by = 0.02),
                       returnType = "data.frame")
  o <- s[!is.na(s$Cc), ]
  w <- o[o$time >= 264 & o$time <= 288, ] # final interval, ~26 half-lives in
  data.frame(dose_mg = md, auc_solved = trap(w$time, w$Cc),
             auc_closed = md / pub_cl)
}))
ss_chk$rel_diff <- abs(ss_chk$auc_solved - ss_chk$auc_closed) / ss_chk$auc_closed

ss_chk |>
  dplyr::rename(
    "Maintenance dose (mg)"            = dose_mg,
    "AUC(0,tau) solved (mg.h/L)"       = auc_solved,
    "Dose/CL (mg.h/L)"                 = auc_closed,
    "Relative difference"              = rel_diff
  ) |>
  knitr::kable(digits = c(0, 4, 4, 10),
               caption = "Steady-state AUC over the 24 h dosing interval vs Dose/CL.")

stopifnot(max(ss_chk$rel_diff) < 1e-3)
```

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the study's actual regimen mix: a 70 mg loading dose for everyone, then 50 mg
q24h in 75% of subjects and 70 mg q24h in 25%, matching the paper's 6:2 split.
Fat-free mass is drawn log-normally, centred on the cohort median of 58.9 kg and
truncated to the observed range of 37.3-81 kg (Table 1).

Cohort size is 200 subjects, the per-arm cap for these vignettes. The paper
simulated 500 for its VPC and 1,000 for its dosing simulations; 200 is ample to
reproduce the published threshold claims and keeps the render fast.

```{r cohort}
# `set.seed()` seeds R's RNG (used for the covariate draw). It does NOT seed
# rxode2's simulation RNG, and rxode2's streams are partitioned PER SOLVER
# THREAD -- so the eta draws below are reproducible on this machine and differ
# on a machine with a different thread count. Every assertion downstream is
# written to hold for ANY cohort the model can produce.
set.seed(11823646)
rxode2::rxSetSeed(11823646)

n_sub <- 200L
ffm_draw <- pmin(pmax(pub_ref * exp(rnorm(n_sub, 0, 0.28)), 37.3), 81)
md_draw <- ifelse(seq_len(n_sub) <= round(n_sub * 6 / 8), 50, 70)

subj <- tibble::tibble(
  id = seq_len(n_sub),
  FFM = ffm_draw,
  md = md_draw,
  regimen = paste0("LD 70 mg, MD ", md_draw, " mg q24h"),
  cohort = "As studied (70 mg LD; 50 mg in 75%, 70 mg in 25%)"
)

# The paper's sampling schedule (Methods, "Study procedures"), applied to the
# day-1 and day-2 intervals, plus a fine grid over day 2 for the NCA.
samp <- c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 6, 8, 12, 24)
obs_times <- sort(unique(c(samp, 24 + samp, seq(24, 48, by = 0.25))))

events <- dplyr::bind_rows(
  subj |>
    tidyr::expand_grid(time = c(0, 24)) |>
    dplyr::mutate(
      amt = ifelse(time == 0, 70, md),
      rate = amt,
      evid = 1L,
      cmt = "central"
    ),
  subj |>
    tidyr::expand_grid(time = obs_times) |>
    dplyr::mutate(amt = NA_real_, rate = NA_real_, evid = 0L, cmt = "central")
) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  dplyr::select(id, time, amt, rate, evid, cmt, FFM, md, regimen, cohort)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
c(subjects = n_sub, md50 = sum(md_draw == 50), md70 = sum(md_draw == 70),
  ffm_median = median(ffm_draw))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod, events = events,
  keep = c("regimen", "cohort", "md", "FFM")
) |>
  as.data.frame()

# `Cc` is the individual prediction (IPRED). The additive residual SD of
# 1.20 mg/L is large relative to late-interval troughs, so residual-perturbed
# observations can go negative; the exposure checks below use `Cc`.
stopifnot(all(sim$Cc[!is.na(sim$Cc)] >= 0))
```

## Replicating Figure 2 (visual predictive check)

Figure 2 of Abdul-Aziz 2025 is a VPC over the single sampled dosing interval.
The panels below show the 5th, 50th and 95th percentiles of the model's
predictions over the day-2 interval, re-referenced to time after dose, split by
the maintenance dose actually received.

```{r figure-2}
sim |>
  dplyr::filter(!is.na(Cc), time >= 24, time <= 48) |>
  dplyr::mutate(tad = time - 24) |>
  dplyr::group_by(regimen, tad) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05), Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~regimen) +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)", y = "Caspofungin concentration (mg/L)",
    title = "Day-2 dosing interval, 5th / 50th / 95th percentiles",
    caption = "Analogue of Figure 2 of Abdul-Aziz 2025 (single-dosing-interval VPC)."
  )
```

# PKNCA validation

Non-compartmental analysis over the day-2 dosing interval (24-48 h), stratified
by the maintenance dose received.

```{r pknca}
# Use only `!is.na(Cc)` -- adding `time > 0` or `Cc > 0` would drop the
# interval-anchoring records PKNCA needs.
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, regimen, cohort)

conc_obj <- PKNCA::PKNCAconc(
  as.data.frame(sim_nca), Cc ~ time | regimen + id,
  concu = "mg/L", timeu = "h"
)

dose_df <- events |>
  dplyr::filter(evid == 1, time == 24) |>
  dplyr::select(id, time, amt, regimen)

dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | regimen + id,
                             doseu = "mg")

intervals <- data.frame(
  start = 24, end = 48,
  cmax = TRUE, tmax = TRUE, cmin = TRUE,
  auclast = TRUE, cav = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_tab <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "cmin", "cav", "auclast", "half.life")) |>
  dplyr::group_by(regimen, PPTESTCD) |>
  dplyr::summarise(
    median = median(PPORRES, na.rm = TRUE),
    q05 = quantile(PPORRES, 0.05, na.rm = TRUE),
    q95 = quantile(PPORRES, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(parameter = nlmixr2lib::ncaParamLabel(PPTESTCD)) |>
  dplyr::select(regimen, parameter, median, q05, q95)

nca_tab |>
  dplyr::rename(
    "Regimen"          = regimen,
    "NCA parameter"    = parameter,
    "Median"           = median,
    "5th percentile"   = q05,
    "95th percentile"  = q95
  ) |>
  knitr::kable(digits = 3,
               caption = "Day-2 (24-48 h) NCA by maintenance dose. Cmax / Cmin / Cav in mg/L, AUC in mg.h/L, Tmax and t-half in h.")
```

## Cross-check: PKNCA against the exact mass-balance AUC

The mass-balance identity gives each subject's day-2 AUC in closed form, so it
can audit PKNCA's trapezoidal result per subject rather than in aggregate. This
also confirms that the amount returned at `time = 24` is the *pre-dose* amount,
which the identity depends on.

```{r pknca-vs-massbalance}
amt_at <- function(t) {
  x <- sim[!is.na(sim$Cc) & sim$time == t, c("id", "central", "cl")]
  x[order(x$id), ]
}
a24 <- amt_at(24)
a48 <- amt_at(48)
stopifnot(identical(a24$id, a48$id), identical(a24$id, subj$id))

exact_auc <- data.frame(
  id = a24$id,
  auc_exact = (a24$central + subj$md - a48$central) / a24$cl
)

pknca_auc <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(id, auc_pknca = PPORRES)

auc_cmp <- dplyr::inner_join(exact_auc, pknca_auc, by = "id")
stopifnot(nrow(auc_cmp) == n_sub) # a lookup that matched nothing must not pass
auc_cmp$rel_diff <- abs(auc_cmp$auc_pknca - auc_cmp$auc_exact) / auc_cmp$auc_exact

c(max_rel_diff = max(auc_cmp$rel_diff),
  median_exact = median(auc_cmp$auc_exact),
  median_pknca = median(auc_cmp$auc_pknca))

# Trapezoidal error on the 0.25 h day-2 grid, per subject. Deterministic given
# the cohort, so a tight bound is correct here.
stopifnot(max(auc_cmp$rel_diff) < 5e-3)
```

## Comparison against the published exposure

The only NCA-style statistic Abdul-Aziz 2025 publishes is in the Discussion:
"the median AUC(0-24) exposure was 115 mg.h/L (range: 81-151)". That is the
observed cohort value, pooled across both maintenance doses, so the comparison
below pools the simulated cohort the same way.

```{r nca_compare}
sim_pooled <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::mutate(cohort = unique(subj$cohort))

published <- tibble::tribble(
  ~cohort,                                                 ~auclast,
  unique(subj$cohort),                                        115.0
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_pooled,
  reference = published,
  by        = "cohort",
  units     = c(auclast = "mg*h/L"),
  tolerance_pct = 20
)

knitr::kable(cmp, caption = "Simulated vs published median AUC(0-24). * differs from reference by >20%.")
attr(cmp, "footnote")
```

The pooled simulated median is about 13% below the published 115 mg.h/L. Two
features of the source explain the gap, and neither is resolvable from the
paper:

* **The sampled interval is not identified.** Sampling ran over "a single
  dosing interval once the patient had been stabilized on ECMO", 0-24 h after
  the start of an infusion, but the paper never states which caspofungin dose
  that interval belonged to. The typical-value AUC(0-24) is 101.3 mg.h/L for
  the day-1 70 mg loading interval, 90.9 mg.h/L at steady state on 50 mg, and
  127.3 mg.h/L at steady state on 70 mg. The published 115 sits inside that
  spread.
* **n = 8.** The published median is the middle of eight observed values, six
  of which came from the 50 mg arm; its sampling error is large.

The two per-arm medians bracket the published value, which is the substantive
check:

```{r bracket}
arm_med <- nca_tab |>
  dplyr::filter(parameter == nlmixr2lib::ncaParamLabel("auclast")) |>
  dplyr::select(regimen, median)
arm_med

# Structural gate on the CENTRE, not the extremes: a mis-transcribed clearance,
# dose or unit moves the whole distribution by tens of percent. The bound is
# set from the paper's own interval ambiguity (115 vs 90.9 = -21%, 115 vs 127.3
# = +11%), not from this run, which realised -13.3%.
pooled_pct <- 100 * (median(sim_pooled$PPORRES) - 115) / 115
stopifnot(abs(pooled_pct) < 25)
# The published median must fall between the two maintenance-dose arms.
stopifnot(min(arm_med$median) < 115, max(arm_med$median) > 115)
```

# Replicating the dosing simulations (Figures 3-5)

This is the paper's main reusable output. Probability of target attainment
(PTA) is "the percentage of patients achieving the target ratio of total drug
AUC(0-24)/MIC ... at 48 hours of therapy" (Methods, "Dosing simulations"), i.e.
over the day-2 interval, with targets from Andes et al.: AUC(0-24)/MIC >= 865
for *C. albicans*, >= 450 for *C. glabrata*, and >= 1185 for
*C. parapsilosis*. A regimen is "optimal" at PTA >= 90%. Simulations use the
paper's grid of lean body weights (40, 50, 60, 80 kg) and MICs (0.004, 0.016,
0.032, 0.064, 0.125, 1 mg/L).

Each subject's day-2 AUC is computed with the exact mass-balance identity
validated above, so the PTA carries no trapezoidal error.

```{r pta}
regimens <- tibble::tribble(
  ~regimen,       ~ld,  ~md,
  "LD 70, MD 50",   70,   50,
  "LD 70, MD 70",   70,   70,
  "LD 100, MD 100", 100, 100,
  "LD 150, MD 100", 150, 100
)
lbw_grid <- c(40, 50, 60, 80)
mic_grid <- c(0.004, 0.016, 0.032, 0.064, 0.125, 1)
targets <- c("C. albicans" = 865, "C. glabrata" = 450, "C. parapsilosis" = 1185)

arms <- tidyr::expand_grid(regimens, lbw = lbw_grid) |>
  dplyr::mutate(arm = dplyr::row_number(),
                id_offset = (arm - 1L) * n_sub)

pta_events <- do.call(dplyr::bind_rows, lapply(seq_len(nrow(arms)), function(k) {
  a <- arms[k, ]
  ids <- a$id_offset + seq_len(n_sub)
  d <- tidyr::expand_grid(tibble::tibble(id = ids), time = c(0, 24)) |>
    dplyr::mutate(amt = ifelse(time == 0, a$ld, a$md), rate = amt,
                  evid = 1L, cmt = "central")
  o <- tidyr::expand_grid(tibble::tibble(id = ids), time = c(24, 48)) |>
    dplyr::mutate(amt = NA_real_, rate = NA_real_, evid = 0L, cmt = "central")
  dplyr::bind_rows(d, o) |>
    dplyr::mutate(FFM = a$lbw, regimen = a$regimen, lbw = a$lbw, md = a$md)
}))
stopifnot(!anyDuplicated(unique(pta_events[, c("id", "time", "evid")])))

pta_sim <- rxode2::rxSolve(mod, events = dplyr::arrange(pta_events, id, time, dplyr::desc(evid)),
                           keep = c("regimen", "lbw", "md")) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc))

subject_auc <- pta_sim |>
  dplyr::filter(time %in% c(24, 48)) |>
  dplyr::select(id, time, central, cl, regimen, lbw, md) |>
  tidyr::pivot_wider(names_from = time, values_from = central,
                     names_prefix = "a") |>
  dplyr::mutate(auc = (a24 + md - a48) / cl)
stopifnot(nrow(subject_auc) == nrow(arms) * n_sub, all(is.finite(subject_auc$auc)))
```

## Day-2 exposure is nearly independent of lean body weight

Before the PTA itself, one structural consequence of the model is worth making
explicit: the fat-free-mass covariate acts on **volume only**, and no covariate
was retained on clearance. Since steady-state exposure is governed by
`Dose / CL`, day-2 AUC is therefore almost flat across lean body weight -- which
is why the paper's *C. glabrata* fractional-target-attainment rows in Table 3
are uniformly "+" from 40 to 80 kg.

```{r lbw-invariance}
lbw_med <- subject_auc |>
  dplyr::group_by(regimen, lbw) |>
  dplyr::summarise(median_auc = median(auc), .groups = "drop")

lbw_med |>
  tidyr::pivot_wider(names_from = lbw, values_from = median_auc,
                     names_prefix = "LBW ") |>
  dplyr::rename("Regimen" = regimen) |>
  knitr::kable(digits = 1,
               caption = "Median day-2 AUC(0-24) (mg.h/L) by regimen and lean body weight.")

# Spread across the 40-80 kg range, per regimen. Had the covariate been placed
# on clearance instead of volume, this would be ~2-fold, not a few percent.
spread <- lbw_med |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(pct_spread = 100 * (max(median_auc) - min(median_auc)) / median(median_auc),
                   .groups = "drop")
spread
stopifnot(max(spread$pct_spread) < 20)
```

## Probability of target attainment

```{r pta-table}
pta <- tidyr::expand_grid(
  subject_auc |> dplyr::distinct(regimen, lbw),
  species = names(targets), mic = mic_grid
) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    pta = {
      a <- subject_auc$auc[subject_auc$regimen == regimen & subject_auc$lbw == lbw]
      stopifnot(length(a) == n_sub)
      100 * mean(a / mic >= targets[[species]])
    }
  ) |>
  dplyr::ungroup()

for (sp in names(targets)) {
  print(
    pta |>
      dplyr::filter(species == sp) |>
      dplyr::select(regimen, lbw, mic, pta) |>
      tidyr::pivot_wider(names_from = mic, values_from = pta, names_prefix = "MIC ") |>
      dplyr::rename("Regimen" = regimen, "LBW (kg)" = lbw) |>
      knitr::kable(digits = 1,
                   caption = paste0("PTA (%) on day 2 for ", sp,
                                    " (AUC(0-24)/MIC target >= ", targets[[sp]], ")."))
  )
}
```

```{r pta-figure}
pta |>
  ggplot(aes(factor(mic), pta, colour = regimen, group = regimen)) +
  geom_hline(yintercept = 90, linetype = "dashed") +
  geom_line() +
  geom_point(size = 1.2) +
  facet_grid(species ~ lbw, labeller = labeller(lbw = function(x) paste0("LBW ", x, " kg"))) +
  labs(
    x = "MIC (mg/L)", y = "PTA (%)", colour = "Regimen",
    title = "Day-2 probability of target attainment",
    caption = "Replicates Figures 3-5 of Abdul-Aziz 2025. Dashed line = the 90% optimality threshold."
  ) +
  theme(legend.position = "bottom")
```

## Verifying the paper's published threshold claims

The paper makes six specific PTA claims. Each is tested below as the paper
states it: at the MIC said to be covered, the PTA must be at or above 90% for
**every** lean body weight; at the next MIC on the grid, it must fall below 90%.

```{r pta-claims}
pta_at <- function(reg, sp, m) {
  v <- pta$pta[pta$regimen == reg & pta$species == sp & pta$mic == m]
  if (length(v) != length(lbw_grid)) {
    stop("no PTA rows for '", reg, "' / ", sp, " at MIC ", m)
  }
  v
}

claims <- tibble::tribble(
  ~claim,                                                                      ~regimen,        ~species,          ~mic_ok, ~mic_fail,
  "C. albicans covered to MIC <= 0.064 (LD 70 / MD 50)",                       "LD 70, MD 50",  "C. albicans",       0.064,     0.125,
  "C. albicans covered to MIC <= 0.064 (LD 70 / MD 70)",                       "LD 70, MD 70",  "C. albicans",       0.064,     0.125,
  "C. glabrata covered to MIC <= 0.125 (LD 70 / MD 50)",                       "LD 70, MD 50",  "C. glabrata",       0.125,     1.000,
  "C. glabrata covered to MIC <= 0.125 (LD 70 / MD 70)",                       "LD 70, MD 70",  "C. glabrata",       0.125,     1.000,
  "C. parapsilosis covered to MIC <= 0.032 (MD 50)",                           "LD 70, MD 50",  "C. parapsilosis",   0.032,     0.064,
  "C. parapsilosis covered to MIC <= 0.064 (MD 70)",                           "LD 70, MD 70",  "C. parapsilosis",   0.064,     0.125
) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    min_pta_at_covered_mic = min(pta_at(regimen, species, mic_ok)),
    max_pta_at_next_mic    = max(pta_at(regimen, species, mic_fail)),
    reproduced             = min_pta_at_covered_mic >= 90 & max_pta_at_next_mic < 90
  ) |>
  dplyr::ungroup()

claims |>
  dplyr::select(claim, mic_ok, min_pta_at_covered_mic, mic_fail, max_pta_at_next_mic, reproduced) |>
  dplyr::rename(
    "Published claim"                  = claim,
    "Covered MIC (mg/L)"               = mic_ok,
    "Min PTA (%) across LBW"           = min_pta_at_covered_mic,
    "Next MIC (mg/L)"                  = mic_fail,
    "Max PTA (%) across LBW"           = max_pta_at_next_mic,
    "Reproduced"                       = reproduced
  ) |>
  # Per-column digits: a scalar would round the MIC column and print 0.064
  # as "0.1" and 0.032 as "0.0".
  knitr::kable(digits = c(0, 3, 1, 3, 1, 0),
               caption = "Abdul-Aziz 2025 Results 'Dosing simulations': all six published PTA threshold claims.")

# Each side of every claim clears the 90% threshold by a wide margin. Across
# two independent renders the worst covered-side minimum was 96.0-96.5% and the
# worst uncovered-side maximum 81.5-82.5%, against a Monte-Carlo standard error
# near 1-3 points at n = 200. Do NOT tighten these to a single run's values.
# The gate can still go red: a mis-transcribed clearance, dose or exponent
# shifts these by tens of points and moves a boundary into the wrong MIC bin.
stopifnot(all(claims$reproduced))
```

All six reproduce. This is a strong validation for a model with only two
structural parameters: the published PTA thresholds are a joint function of the
clearance point estimate, its between-subject variability, the dosing regimen
and the accumulation implied by the volume, and getting all six boundaries in
the right MIC bin exercises every one of them.

The paper's remaining dosing claims are **fractional** target attainment (FTA,
Table 3), which weights the PTA by published MIC *distributions* for each
*Candida* species (the paper's references 19 and 20). Those distributions are
not reproduced anywhere in the paper, so FTA cannot be recomputed from the
source on disk and is not attempted here.

# Assumptions and deviations

* **Between-subject variability scale.** Table 2 reports the BSV rows as
  percentages ("CL (%) 25.5", "V (%) 7.41"). Read as coefficients of variation
  -- Monolix's standard summary for a parameter declared log-normal, and
  consistent with the Methods' explicit exponential / log-normal statement --
  they back-transform to `omega^2 = log(1 + CV^2)`, giving 0.062996 and
  0.0054758. The alternative reading, that the printed percentages are raw
  omega SDs scaled by 100, gives `omega_CL` 0.255 vs 0.251 and `omega_V` 0.0741
  vs 0.0740: under 2% relative difference, which no validation gate here or in
  the paper can resolve. The CV reading is used.
* **Both etas are weakly identified.** BSV on CL has %RSE 34.3 and BSV on V has
  %RSE 101 with a bootstrap 95% CI of 1.03-22.6% -- essentially unidentified in
  a cohort of 8. Treat the stochastic layer as indicative; the typical-value
  predictions are far better supported.
* **The fat-free-mass exponent is poorly identified.** 1.24 carries a bootstrap
  95% CI of 0.08-1.91, which spans everything from no scaling to strongly
  supra-linear. It is encoded as the estimated point estimate because that is
  what the paper fitted and reported, but a user extrapolating far outside the
  observed 37.3-81 kg range should be aware the exponent is barely
  distinguishable from 1 (the theory-based value for a volume).
* **Fat-free mass distribution in the virtual cohort.** The paper reports only
  the median (58.9 kg) and range (37.3-81) of lean body weight, not a
  distribution. The cohort draws log-normally around the median with
  `sigma = 0.28` on the log scale, truncated to the observed range; that sigma
  was chosen so the draw spans the reported range, not fitted to anything.
  Height and sex, which the Janmahasatian equation needs, are not reported per
  subject, so `FFM` is drawn directly rather than derived.
* **The sampled dosing interval is not identified in the source.** Sampling
  covered "a single dosing interval once the patient had been stabilized on
  ECMO" but the paper never says which caspofungin dose. Day 2 is used
  throughout here because that is the interval the paper's own PTA analysis
  targets ("at 48 hours of therapy"). See the exposure comparison above for how
  this affects the published 115 mg.h/L median.
* **ECMO is not a covariate and cannot be one in this model.** All 8 patients
  were on ECMO, so the treatment indicator is constant. The model therefore
  describes caspofungin PK *in* an all-ECMO population; it does not quantify an
  ECMO-vs-no-ECMO contrast, and the paper is explicit that "the true influence
  may not be able to be fully quantified without non-ECMO controls for
  comparison". Users wanting an explicit ECMO effect should look to
  `Valadez_2025_cefepime.R` or `Kang_2020_cefpirome.R`, which fit
  `ECMO_STATUS` directly.
* **ECMO and renal replacement therapy are confounded here.** 5 of 8 patients
  were on concomitant RRT, and neither RRT nor any ECMO variable was retained.
  With n = 8 the two extracorporeal circuits cannot be separated.
* **Screened-but-unretained covariates.** Age, sex, ideal and adjusted body
  weight, total body weight, BMI, creatinine clearance, serum albumin,
  bilirubin, blood urea nitrogen, APACHE II and SOFA scores, receipt of RRT,
  and ECMO duration / mode / flow rate were all screened (Methods, "Covariate
  screening and model development"); only fat-free mass on volume survived.
  Those with an existing canonical column are recorded in the model file's
  `covariatesDataExcluded` metadata. Adjusted body weight, the SOFA score, and
  the ECMO treatment variables are described there in prose only -- no
  canonical column exists for them, and registering one for a covariate that no
  model uses would be an unwarranted addition to the register.
* **Supplementary material not obtainable.** The paper states that "the
  pharmacokinetic model-building process is further detailed in Supplementary
  materials". The ASM supplement endpoint returns HTTP 403, and the EuropePMC
  `supplementaryFiles` archive for PMC11823646 contains only per-figure and
  per-equation raster images, not the supplement document. Nothing in the model
  depends on it: all six final parameter estimates are in Table 2, and the
  covariate equation was recovered from the typeset equation image
  (`aac.01435-24.m002.jpg`) and confirmed against `pdftotext -layout` output.
  The unobtainable content is the model-building progression (OFV / BICc step
  table), which is provenance for model *selection* rather than any value used
  here.
* **Fractional target attainment is not reproduced**, because the *Candida* MIC
  distributions it weights by are cited rather than printed. See the end of the
  dosing-simulation section.
* **Residual error and simulated observations.** The additive residual SD of
  1.20 mg/L is large relative to late-interval troughs (typical day-2 Cmin is
  around 1.7 mg/L), so residual-perturbed observations can be negative. All
  exposure and PTA calculations here use the model's `Cc` prediction (IPRED),
  which is the quantity the paper's own AUC-based targets are defined on.
