---
title: "Apomorphine (Agbo 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Apomorphine (Agbo 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Agbo_2021_apomorphine"))$reference`
* Description: `r rxode2::rxode(readModelDb("Agbo_2021_apomorphine"))$description`
* Article: <https://doi.org/10.1111/cts.13008> (open access)

Agbo et al. fitted apomorphine and its inactive metabolite apomorphine sulfate
simultaneously after apomorphine sublingual film (APL, KYNMOBI) and two
subcutaneous apomorphine products (APOKYN, APO-go). The structure (Figure 1 of
the paper) is:

* **Sublingual route**: the film dose is split. A fraction `Biosl` (not
  swallowed) enters the sublingual chain `depot -> transit1 -> transit2 ->
  central` (paper CMT 1 -> 7 -> 8 -> 2), all three steps at the same rate
  constant `ka` (k17 = k78 = k82), with bioavailability relative to the
  subcutaneous route `Biorsc`. The swallowed remainder `F5 = 1 - Biosl` is
  absorbed from the gastrointestinal tract (`depot_sulf`, paper CMT 5)
  **directly as apomorphine sulfate** at rate `ka_sulf` (k53).
* **Subcutaneous route**: `depot2 -> transit3 -> transit4 -> central` (paper
  CMT 4 -> 9 -> 10 -> 2), all at `ka2` (k49 = k9T10 = k10T2), `F4 = 1`.
* **Apomorphine**: two compartments (central, peripheral1; k26 and k62). All
  apomorphine elimination is first-order metabolism to the sulfate at
  `k23 = CL/V2`.
* **Apomorphine sulfate**: one compartment (`central_sulf`, V3/F) with
  first-order elimination `k30`.

## Dosing the model

A sublingual film dose of `X` mg is **two dose records** at the same time: `amt
= X` into `depot` and `amt = X` into `depot_sulf`. The model applies
`f(depot) = Biosl * Biorsc * (DOSE_APOMORPHINE_SL_MG / 20)^-0.206` and
`f(depot_sulf) = 1 - Biosl`, so the two records together carry the whole dose.
A subcutaneous dose is one record into `depot2`. The model has two error
endpoints (`Cc` for apomorphine, `Cc_sulf` for the sulfate, both ng/mL), so
observation rows carry `dvid` rather than `cmt`; both concentrations come back
as columns of every solve.

# Population

The analysis pooled 9 studies (5 phase I, 3 phase II, 1 phase III;
Supplementary Table S1) with 158 individuals: 87 healthy subjects and 71
patients with Parkinson's disease (PD) and OFF episodes. There were 2485
apomorphine samples from all 158 and 1182 apomorphine sulfate samples from 101
(Results, "Model population"). Healthy subjects were younger (mean 27.2 vs 65.0
years), lighter (mean 65.8 vs 80.0 kg), predominantly Asian (98%) where
patients were predominantly White (94%), and had higher creatinine clearance
(mean 114 vs 94 mL/min; Table 2). Overall, 44 of 158 (27.8%) were female.
Sublingual film doses were 10-35 mg and 50 mg; subcutaneous doses were 2-5 mg.

The same information is available programmatically via
`readModelDb("Agbo_2021_apomorphine")()$population`.

# Source trace

Every value is from Table 3 of Agbo 2021 unless noted. Methods Eq. 2 defines
`%CV = sqrt(omega^2) * 100`, so each variance is `(CV / 100)^2`.

| Parameter | Value | Source location |
|-----------|-------|-----------------|
| `lcl` (CL/F) | log(80.7 L/h) | Table 3 'CL/F' |
| `lvc` (V2/F) | log(438 L) | Table 3 'V2/F' |
| `lk12` (k26) | log(0.613 1/h) | Table 3 'k26' |
| `lk21` (k62) | log(0.00480 1/h) | Table 3 'k62' |
| `lka` (sublingual ka) | log(6.58 1/h) | Table 3 'ka for sublingual administration' |
| `lka2` (subcutaneous ka) | log(17.6 1/h) | Table 3 'ka for subcutaneous administration' |
| `lfdepot` (Biorsc) | log(0.202) | Table 3 'Fraction absorbed relative to subcutaneous administration' |
| `lfdepot_sulf` (F5) | log(1 - 0.910) | Table 3 'Fraction not swallowed ...' = 0.910; Figure 1 `F5 = 1 - Biosl` |
| `lvc_sulf` (V3/F) | log(1.42 L) | Table 3 'V3/F' |
| `lkel_sulf` (k30) | log(1.28 1/h) | Table 3 'k30' |
| `lka_sulf` (k53) | log(0.205 1/h) | Table 3 'ka for apomorphine sulfate absorption from the gastrointestinal tract' |
| `e_dose_fdepot` | -0.206 (reference 20 mg) | Table 3; reference in Results 'Final model' |
| `e_slct_ka` | -0.194 (reference 2 min) | Table 3; reference in Results 'Final model' |
| `e_wt_vc` | 1.53 (reference 69.3 kg) | Table 3; reference in Results 'Final model' |
| `e_study_cth103_vc` | -0.555, form `1 + theta * X` | Table 3; Methods Eq. 5; Results 'proportional shift' |
| `e_sexf_vc_sulf` | -0.310, form `1 + theta * X` | Table 3; Methods Eq. 5; Results 'proportional shift' |
| `etalcl` (paper: IIV on k23) | 0.359^2 | Table 3 IIV 'k23' 35.9 %CV |
| `etalvc` | 0.417^2 | Table 3 IIV 'V2/F' 41.7 %CV |
| `etalvc_sulf` | 0.403^2 | Table 3 IIV 'V3/F' 40.3 %CV |
| `etalka` | 0.357^2 | Table 3 IIV 'ka for sublingual administration' 35.7 %CV |
| `etalfdepot` (IIV on F1) | 0.383^2 | Table 3 IIV 'F1' 38.3 %CV |
| `propSd` | 0.443 | Table 3 residual 'Apomorphine' 44.3% |
| `propSd_sulf` | 0.580 | Table 3 residual 'Apomorphine sulfate' 58.0% |
| Compartment structure and rate-constant equalities | n/a | Figure 1 and its caption |
| `k23 = CL/V2` | n/a | Table 3 abbreviations: 'k23 ... equal to CL/V2' |

# Typical-value checks

`zeroRe()` removes all between-subject and residual variability. The helper
below builds a one-subject event table for either route; `n_dose` doses are
given every 2 hours.

```{r helpers}
mod <- readModelDb("Agbo_2021_apomorphine")
mod_typ <- rxode2::zeroRe(mod)

obs_grid <- sort(unique(c(seq(0, 2, by = 0.01), seq(2, 24, by = 0.05))))

make_events <- function(id, route, dose, wt, ct = 3, sexf = 0, n_dose = 1,
                        treatment = NA_character_) {
  dose_times <- (seq_len(n_dose) - 1) * 2
  if (route == "SL") {
    doses <- dplyr::bind_rows(
      data.frame(time = dose_times, amt = dose, cmt = "depot"),
      data.frame(time = dose_times, amt = dose, cmt = "depot_sulf")
    )
  } else {
    doses <- data.frame(time = dose_times, amt = dose, cmt = "depot2")
  }
  doses$evid <- 1L
  doses$dvid <- NA_integer_
  obs <- data.frame(
    time = obs_grid, amt = NA_real_, cmt = NA_character_,
    evid = 0L, dvid = 1L
  )
  dplyr::bind_rows(doses, obs) |>
    dplyr::mutate(
      id = id,
      WT = wt,
      SEXF = sexf,
      STUDY_CTH103 = 0,
      # Only f(depot) reads this column, so its value on subcutaneous
      # records is irrelevant; the reference 20 mg is used there.
      DOSE_APOMORPHINE_SL_MG = if (route == "SL") dose else 20,
      SL_CONTACT_TIME_MIN = ct,
      treatment = treatment
    ) |>
    dplyr::relocate(id, time, amt, evid, cmt, dvid) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

solve_typ <- function(events) {
  sim <- rxode2::rxSolve(
    mod_typ,
    events = events,
    keep = c("treatment"),
    returnType = "data.frame",
    useLinCmt = FALSE,
    atol = 1e-10,
    rtol = 1e-10
  )
  # rxSolve omits the id column for a one-subject solve.
  if (!"id" %in% names(sim)) {
    sim$id <- unique(events$id)
  }
  sim
}

# PKNCA over 0-24 h for one analyte column of a solve.
nca_0_24 <- function(sim, events, conc_col) {
  conc <- sim |>
    dplyr::transmute(id, treatment, time, conc = .data[[conc_col]]) |>
    dplyr::filter(!is.na(conc))
  conc <- dplyr::bind_rows(
    conc,
    conc |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, conc = 0)
  ) |>
    dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
    dplyr::arrange(id, treatment, time)
  dose_df <- events |>
    dplyr::filter(evid == 1, cmt != "depot_sulf") |>
    dplyr::select(id, time, amt, treatment)
  conc_obj <- PKNCA::PKNCAconc(conc, conc ~ time | treatment + id)
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)
  intervals <- data.frame(start = 0, end = 24, cmax = TRUE, auclast = TRUE)
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}

nca_wide <- function(res) {
  as.data.frame(res) |>
    dplyr::select(treatment, PPTESTCD, PPORRES) |>
    dplyr::group_by(treatment, PPTESTCD) |>
    dplyr::summarise(value = stats::median(PPORRES), .groups = "drop") |>
    tidyr::pivot_wider(names_from = PPTESTCD, values_from = value)
}
```

## Figure 2: the typical patient with PD, and body weight

Figure 2 and the Results text describe a typical patient with PD: 78 kg male,
a single 20 mg sublingual film dose, 3 min contact time under the tongue. For
that patient the paper reports Cmax 4.48 ng/mL and AUC 10.60 ng*h/mL, and for
a 114 kg patient Cmax 2.62 ng/mL and AUC 6.77 ng*h/mL.

```{r figure-2}
ev_fig2 <- dplyr::bind_rows(
  make_events(1L, "SL", 20, wt = 78, treatment = "78 kg"),
  make_events(2L, "SL", 20, wt = 114, treatment = "114 kg")
)
sim_fig2 <- solve_typ(ev_fig2)
nca_fig2 <- nca_0_24(sim_fig2, ev_fig2, "Cc")

published_fig2 <- tibble::tribble(
  ~treatment, ~cmax, ~auclast,
  "78 kg",     4.48,  10.60,
  "114 kg",    2.62,   6.77
)
cmp_fig2 <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_fig2,
  reference = published_fig2,
  by = "treatment",
  units = c(cmax = "ng/mL", auclast = "ng*h/mL"),
  tolerance_pct = 20
)
knitr::kable(
  cmp_fig2,
  caption = paste(
    "Typical patient with PD, single 20 mg sublingual dose.",
    "Reference AUC is the paper's 'AUC0-inf'; simulated is PKNCA AUC0-24",
    "(see Assumptions)."
  )
)
```

```{r figure-2-gate}
w2 <- nca_wide(nca_fig2)
cmax_78 <- w2$cmax[w2$treatment == "78 kg"]
cmax_114 <- w2$cmax[w2$treatment == "114 kg"]
auc_78 <- w2$auclast[w2$treatment == "78 kg"]
auc_114 <- w2$auclast[w2$treatment == "114 kg"]
c(
  cmax_78 = cmax_78, cmax_114 = cmax_114,
  auc_ratio_114_vs_78 = auc_114 / auc_78
)
stopifnot(
  # Deterministic typical-value solve: Cmax reproduces the text to about 1%.
  abs(cmax_78 / 4.48 - 1) < 0.05,
  abs(cmax_114 / 2.62 - 1) < 0.05,
  # Body-weight effect on AUC: text ratio 6.77 / 10.60 = 0.639, Figure 2
  # forest plot 95th-percentile point about 0.61; simulated 0.618.
  abs((auc_114 / auc_78) / (6.77 / 10.60) - 1) < 0.06
)
```

The body-weight effect acts only on V2/F. Because apparent clearance does not
depend on weight, a heavier patient has a larger V2/F and therefore a smaller
`k23 = CL/V2` and a lower Cmax and early exposure. This reading reproduces
both Cmax values to about 1%. The alternative, in which `k23` itself is the
weight-independent parameter so that clearance grows with V2/F, is 3-5% worse
on every Figure 2 and Figure 3 comparison (see Assumptions).

## Dose dependence of sublingual bioavailability

The Discussion states that the relative bioavailability of a 30 mg sublingual
dose is about 80% of a 10 mg dose; the model gives `3^-0.206`.

```{r dose-proportionality}
ev_dp <- dplyr::bind_rows(
  make_events(1L, "SL", 10, wt = 78, treatment = "10 mg"),
  make_events(2L, "SL", 30, wt = 78, treatment = "30 mg")
)
w_dp <- nca_wide(nca_0_24(solve_typ(ev_dp), ev_dp, "Cc"))
rel_f <- (w_dp$auclast[w_dp$treatment == "30 mg"] / 30) /
  (w_dp$auclast[w_dp$treatment == "10 mg"] / 10)
rel_f
stopifnot(abs(rel_f - 0.80) < 0.01)
```

## Figure 3: maximal dosing, typical patient

Figure 3 and the Results give the median predicted apomorphine exposure for
1000 simulated patients with PD after 30 or 35 mg sublingual film, or 5 or 6 mg
subcutaneous apomorphine, every 2 hours for 5 doses: AUC0-24 66.7, 75.4, 66.7
and 80.0 ng*h/mL and Cmax about 9, 10, 10 and 12 ng/mL. The typical 78 kg
patient reproduces these medians closely.

```{r figure-3-typical}
regimens <- tibble::tribble(
  ~treatment,     ~route, ~dose,
  "APL 30 mg",    "SL",   30,
  "APL 35 mg",    "SL",   35,
  "SC 5 mg",      "SC",    5,
  "SC 6 mg",      "SC",    6
)
ev_fig3_typ <- dplyr::bind_rows(lapply(seq_len(nrow(regimens)), function(i) {
  make_events(
    i, regimens$route[i], regimens$dose[i],
    wt = 78, n_dose = 5, treatment = regimens$treatment[i]
  )
}))
sim_fig3_typ <- solve_typ(ev_fig3_typ)
nca_fig3_typ <- nca_0_24(sim_fig3_typ, ev_fig3_typ, "Cc")

published_fig3 <- tibble::tribble(
  ~treatment,  ~cmax, ~auclast,
  "APL 30 mg",    9,     66.7,
  "APL 35 mg",   10,     75.4,
  "SC 5 mg",     10,     66.7,
  "SC 6 mg",     12,     80.0
)
knitr::kable(
  nlmixr2lib::ncaComparisonTable(
    simulated = nca_fig3_typ,
    reference = published_fig3,
    by = "treatment",
    units = c(cmax = "ng/mL", auclast = "ng*h/mL"),
    tolerance_pct = 20
  ),
  caption = paste(
    "Typical 78 kg patient with PD vs the paper's cohort medians",
    "(Figure 3; Results). Published Cmax values are the text's rounded",
    "approximations."
  )
)

w3 <- nca_wide(nca_fig3_typ) |>
  dplyr::inner_join(published_fig3, by = "treatment", suffix = c("", "_pub"))
stopifnot(
  all(abs(w3$auclast / w3$auclast_pub - 1) < 0.05),
  all(abs(w3$cmax / w3$cmax_pub - 1) < 0.10)
)
```

The paper also states that subcutaneous dosing gives apomorphine sulfate Cmax
about 35% of that after sublingual dosing at the two comparative dose pairs.
The sublingual excess comes from the swallowed fraction absorbed directly as
the sulfate.

```{r sulfate-ratio}
nca_sulf <- nca_wide(nca_0_24(sim_fig3_typ, ev_fig3_typ, "Cc_sulf"))
cmax_s <- stats::setNames(nca_sulf$cmax, nca_sulf$treatment)
sulf_ratio <- c(
  "SC 5 mg / APL 30 mg" = cmax_s[["SC 5 mg"]] / cmax_s[["APL 30 mg"]],
  "SC 6 mg / APL 35 mg" = cmax_s[["SC 6 mg"]] / cmax_s[["APL 35 mg"]]
)
sulf_ratio
stopifnot(all(abs(sulf_ratio - 0.35) < 0.05))
```

# Virtual cohort

The paper simulated 1000 patients with PD by bootstrapping complete covariate
vectors from the observed PD data. Those data are not public. The cohort below
draws 200 patients per regimen. Weight is log-normal around the Table 2 PD
mean (80 kg), with an assumed 20% CV, and draws are rejected and redrawn
outside the observed PD range (50.7-146.1 kg). 31% are female (Table 2), all
are outside study CTH-103, and every film has 3 min contact time (the Figure 2
typical value).

```{r cohort}
set.seed(20210401)
n_per_arm <- 200L

draw_wt <- function(n) {
  out <- numeric(0)
  while (length(out) < n) {
    x <- stats::rlnorm(n, meanlog = log(80) - 0.2^2 / 2, sdlog = 0.2)
    out <- c(out, x[x >= 50.7 & x <= 146.1])
  }
  out[seq_len(n)]
}

cohort_events <- dplyr::bind_rows(lapply(seq_len(nrow(regimens)), function(i) {
  wt <- draw_wt(n_per_arm)
  sexf <- stats::rbinom(n_per_arm, 1, 0.31)
  dplyr::bind_rows(lapply(seq_len(n_per_arm), function(j) {
    make_events(
      (i - 1L) * n_per_arm + j, regimens$route[i], regimens$dose[i],
      wt = wt[j], sexf = sexf[j], n_dose = 5,
      treatment = regimens$treatment[i]
    )
  }))
}))
stopifnot(
  !anyDuplicated(unique(cohort_events[, c("id", "time", "evid", "cmt")])),
  dplyr::n_distinct(cohort_events$id) == 4L * n_per_arm
)
```

# Simulation

```{r simulate}
rxode2::rxSetSeed(20210401)
sim <- rxode2::rxSolve(
  mod,
  events = cohort_events,
  keep = c("treatment", "WT"),
  returnType = "data.frame",
  useLinCmt = FALSE
)
```

# Replicate published figures

```{r figure-3, fig.width = 8, fig.height = 6}
# Replicates Figure 3 of Agbo 2021: median and 90% prediction interval of
# apomorphine concentration, 5 doses every 2 h. Uses the individual
# prediction (no residual error), as the paper's exposure summaries do.
sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(treatment, time) |>
  dplyr::summarise(
    Q05 = stats::quantile(Cc, 0.05),
    Q50 = stats::quantile(Cc, 0.50),
    Q95 = stats::quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_line(aes(y = Q05), linetype = "dotted") +
  geom_line(aes(y = Q95), linetype = "dotted") +
  geom_line() +
  facet_wrap(~treatment) +
  labs(
    x = "Time since first dose (h)",
    y = "Apomorphine concentration (ng/mL)",
    caption = "Replicates Figure 3 of Agbo 2021 (virtual cohort, 200 per regimen)."
  )
```

```{r figure-3-sulfate, fig.width = 8, fig.height = 6}
# Apomorphine sulfate under the same regimens (not plotted in the paper).
sim |>
  dplyr::filter(!is.na(Cc_sulf)) |>
  dplyr::group_by(treatment, time) |>
  dplyr::summarise(
    Q05 = stats::quantile(Cc_sulf, 0.05),
    Q50 = stats::quantile(Cc_sulf, 0.50),
    Q95 = stats::quantile(Cc_sulf, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~treatment) +
  labs(
    x = "Time since first dose (h)",
    y = "Apomorphine sulfate concentration (ng/mL)",
    caption = "Median and 90% prediction interval, virtual cohort."
  )
```

# PKNCA validation

```{r pknca}
nca_cohort <- nca_0_24(sim, cohort_events, "Cc")
cmp_cohort <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_cohort,
  reference = published_fig3,
  by = "treatment",
  units = c(cmax = "ng/mL", auclast = "ng*h/mL"),
  tolerance_pct = 20
)
knitr::kable(
  cmp_cohort,
  caption = paste(
    "Virtual-cohort medians (200 per regimen) vs the paper's medians of",
    "1000 bootstrapped patients with PD (Figure 3; Results). * differs by >20%."
  )
)

wc <- nca_wide(nca_cohort) |>
  dplyr::inner_join(published_fig3, by = "treatment", suffix = c("", "_pub"))
stopifnot(
  # Centre only. The cohort's weight distribution is an assumption and each
  # median is over 200 subjects whose draw differs across rxode2 builds and
  # thread counts, so this is a coarse envelope; the structural check is the
  # deterministic typical-value gate above (within 1% on AUC0-24). Observed
  # while authoring: AUC0-24 -11.4% to +3.5%, Cmax -11.4% to +10.7%. A
  # mis-transcribed clearance, volume, bioavailability or unit moves these
  # by well over 25%.
  all(abs(wc$auclast / wc$auclast_pub - 1) < 0.25),
  all(abs(wc$cmax / wc$cmax_pub - 1) < 0.25)
)
```

The paper's Table 4 gives individual predicted exposure after a single 10 mg
film in the patients with PD: median Cmax 2.45 and 2.07 ng/mL and median AUC
5.34 and 5.31 ng*h/mL for normal renal function and mild renal impairment.
The typical 78 kg patient with PD sits close to these medians:

```{r table-4}
ev_t4 <- make_events(1L, "SL", 10, wt = 78, treatment = "10 mg")
w_t4 <- nca_wide(nca_0_24(solve_typ(ev_t4), ev_t4, "Cc"))
w_t4
```

# Assumptions and deviations

- **`k23 = CL/V2` with a weight-independent CL.** Table 3 reports CL/F as the
  estimated parameter and puts the IIV on `k23`, "equal to CL/V2". The model
  therefore computes `kel = cl / vc` from the individual V2/F, and the `k23`
  random effect is carried as `etalcl`. Under `k23 = CL/V2 * exp(eta)` that is
  exactly an eta on CL. The maintainers chose this over a `k23` that is itself
  weight-independent, in which clearance would grow with V2/F. This reading
  reproduces the Figure 3 AUC0-24 medians within 1% for a typical 78 kg patient
  and the Figure 2 Cmax values within about 1%; the alternative is 3-5% off on
  every one. It also agrees with the Discussion, which says apparent clearance
  is not influenced by weight. The control stream is not published, so it
  cannot be confirmed whether V2 in the `k23` denominator included the V2/F
  random effect. The typical-value checks cannot tell the two apart.
- **"AUC0-inf" in the paper is a finite-window value.** With `k62 = 0.0048/h`,
  about three quarters of the true AUC0-inf lies in a terminal phase with a
  half-life of several hundred hours at concentrations near 0.03 ng/mL. The
  paper's "AUC0-inf" of 10.60 ng*h/mL for the typical patient is close to the
  model's AUC0-24 (9.9 ng*h/mL) and far from the model's true AUC0-inf (about
  45 ng*h/mL). This vignette therefore compares against PKNCA AUC0-24. The
  paper's exact NCA window and extrapolation are not reported; the remaining
  7-10% gap on these two AUCs is not otherwise explained. Cmax, the body-weight
  AUC ratio and all Figure 3 values agree.
- **Residual error.** Methods Eq. 3 introduces a log-transformed error model;
  the Results state that a proportional residual error model was used for both
  analytes. Both are encoded as proportional (`prop()`) with SD 0.443 and
  0.580, the Table 3 percentages read as fractions.
- **IIV on F1** multiplies the whole sublingual bioavailability
  `Biosl * Biorsc * dose effect` (log-normal, Methods Eq. 1). The swallowed
  fraction `F5 = 1 - Biosl` carries no IIV, because Table 3 lists none for
  it.
- **Categorical covariates** (study CTH-103 on V2/F, female sex on V3/F) use
  the `1 + theta * X` form of Methods Eq. 5. The Results describe both as "the
  proportional shift from the reference condition", and the negative estimates
  match Eq. 5's constraint that `theta > -1`.
- **Apomorphine sulfate units.** Amounts move 1:1 from apomorphine to the
  sulfate compartment, with no molecular-weight conversion. V3/F (1.42 L) is
  an apparent volume estimated against the sulfate data, so `Cc_sulf` is in the
  units of the fitted observations (ng/mL, as for all concentrations in the
  paper). The supplementary goodness-of-fit and VPC figures (S1-S3), which
  would show the sulfate concentration scale, were not included in the open
  access deposit. The published SC/sublingual sulfate Cmax ratio (about 35%)
  is reproduced.
- **Contact time and dose covariates are per administration.** The
  sublingual-only covariates `SL_CONTACT_TIME_MIN` and
  `DOSE_APOMORPHINE_SL_MG` must be present on every row but only affect the
  sublingual chain. Their value on subcutaneous records is irrelevant.
- **Virtual cohort.** The PD weight distribution (log-normal, mean 80 kg, 20%
  CV, truncated to the observed range) is an assumption, because the paper
  bootstrapped its own PD covariate data. All subjects use a 3 min contact time
  and are outside study CTH-103. The Figure 3 comparison is gated on medians
  only.
- **Not reproduced.** The Figure 2 forest plot's dose rows come from a
  simulation with parameter uncertainty and do not equal `(dose/20)^-0.206`
  exactly (for example 1.18 at 10 mg vs 1.15 from the point estimate). The
  point estimate is used here. The Discussion's 30-vs-10 mg claim (80%) is
  reproduced exactly.
- No erratum or correction notice was found for this article (EuropePMC,
  checked 2026-09-28).
