---
title: "Darbepoetin alfa in chemotherapy-induced anemia (Agoram 2006)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Darbepoetin alfa in chemotherapy-induced anemia (Agoram 2006)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Agoram 2006 develops **two** models, fitted sequentially, and nlmixr2lib ships
them as two files so that each is usable on its own:

| Model | Role | Source table |
|---|---|---|
| `Agoram_2006_darbepoetin_alfa_cia` | Population PK of darbepoetin alfa in chemotherapy-induced anemia (CIA), freely estimated with its full random-effects structure | Table 3 |
| `Agoram_2006_darbepoetin_alfa_cia_hemoglobin` | PK/PD model of the hemoglobin response, with the PK layer held at the Table 3 fixed effects | Table 4 |

The paper is explicit that the second stage reuses the first: "the Pk model
consisted of fixed-effects parameters obtained from the full covariate model;
random IIV and RRV in the Pk model were ignored". The PD file therefore wraps
every PK entry in `fixed()` and carries no PK etas -- that is the model the
authors actually fitted, not a simplification introduced here.

* Citation: `r rxode2::rxode(readModelDb("Agoram_2006_darbepoetin_alfa_cia"))$reference`
* Article: <https://doi.org/10.1208/aapsj080364>
* PubMed Central: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2761062/>

```{r load-models}
mod_pk <- readModelDb("Agoram_2006_darbepoetin_alfa_cia")
mod_pd <- readModelDb("Agoram_2006_darbepoetin_alfa_cia_hemoglobin")

ui_pk <- rxode2::rxode(mod_pk)
ui_pd <- rxode2::rxode(mod_pd)
```

**PK model** -- `r ui_pk$description`

**PK/PD model** -- `r ui_pd$description`

Note that a *different* Agoram 2006 paper is also in the library:
`Agoram_2006_darbepoetin_alfa` is the healthy-subject popPK model from
*Br J Clin Pharmacol* 63(1):41-52. Same first author, same year, same drug,
different study population -- hence the `_cia` suffix here.

# Population

The PK model was built on 140 patients and 1442 serum darbepoetin alfa
concentrations pooled from two Amgen studies (990146, n = 56, IV or SC
2.25 ug/kg QW; 20010162, n = 84, SC 6.75 ug/kg Q3W). The PK/PD model was built
on a larger, partly disjoint set of 573 patients and 6356 hemoglobin
observations from studies 20010162, 980290 and 980291, spanning SC doses from
0.5 ug/kg QW to 15 ug/kg Q3W. Hemoglobin data from study 990146 were excluded
from the PD analysis because its hemoglobin entry criterion differed
substantially from the other studies (Agoram 2006 Table 1).

Patients were adults (>= 18 years) with nonmyeloid malignancies receiving
cyclic chemotherapy, anemic as a result of cancer or chemotherapy, with ECOG
performance status 0-2 and adequate renal and hepatic function. Baseline
characteristics (Agoram 2006 Table 2) for the combined PK/PD set: 70% female,
age mean 60.1 years (SD 12.3, range 20-91), body weight mean 69.9 kg (SD 16.2,
range 39-136), baseline hemoglobin mean 9.96 g/dL (SD 1.00, range 5.5-12.3).
Tumor types were breast 31%, lung 17%, gastrointestinal 17%, gynecologic 17%,
genitourinary 7%, lymphoma 2%, other 9%. Race was screened as a covariate but
is not tabulated in the publication.

The same information is available programmatically from either model's
`population` metadata:

```{r population}
str(ui_pd$population, max.level = 1, give.attr = FALSE)
```

# Source trace

Every `ini()` entry in both model files carries an in-file comment naming the
table it came from. They are collected here for review.

## Structure

| Equation / structure | Source location |
|---|---|
| Two-compartment disposition, first-order SC absorption, bioavailability `F` | Figure 1A; Results, "Pharmacokinetic Model" |
| Total measured serum concentration = darbepoetin alfa + endogenous EPO (assay cross-reactivity) | Analytical Methods; Equation 2 |
| Log-normal IIV, `P_i = TVP * exp(eta)` | Equation 1 |
| Log-transformed exponential (log-normal) PK residual error | Equation 2 |
| Normalized power / categorical covariate model | Equation 3 |
| `TVCL = theta_CL * theta_1^X * (BWT/70)^theta_2`, `X = 1` if PCNT > 2 | Equation 12 |
| `TVV1 = theta_V1 * (BWT/70)^theta_3` | Equation 13 |
| Indirect response, `dHb/dt = Rin - Kout * Hb`, `Hb(0) = Hb0`, `0 = Rin - Kout * Hb0` | Equations 4, 5, 6 |
| `dA1/dt = Rin * Stim - kPT * A1` (progenitor compartment) | Equation 7 |
| `Stim = 1 + Smax * C_DA / (S50 + C_DA)` | Equation 8 |
| `dA2/dt = kPT * A1 - kLS * A2`; `dAi/dt = kLS * A(i-1) - kLS * Ai`, `i = 3,4,5` | Equations 9, 10 |
| `Hb = sum(Ai, i = 2..5)` | Equation 11 |
| `kPT = 1 / RBCPT`, `kLS = 4 / RBCLS` | Text following Equation 6 |
| `TVS50 = theta_S50 * theta_4^X`, `X = 1` if PCNT > 0 | Equation 14 |

## Parameters

| Parameter | Model file entry | Value | Source |
|---|---|---|---|
| CL | `lcl` | 2010 mL/day | Table 3 |
| V1 | `lvc` | 3390 mL | Table 3 |
| V2 | `lvp` | 251 mL | Table 3 |
| Q | `lq` | 2900 mL/day | Table 3 |
| Ka | `lka` | 0.318 /day | Table 3 |
| F | `lfdepot` | 0.443 | Table 3 |
| C0 (eEPO) | `leepo` | 0.415 ng/mL | Table 3 |
| theta_1 (PCNT on CL) | `e_conmed_platin_gt2_cl` | 0.737 | Table 3 |
| theta_2 (BWT on CL) | `e_wt_cl` | 0.623 | Table 3 |
| theta_3 (BWT on V1) | `e_wt_vc` | 0.639 | Table 3 |
| omega^2 CL, CL-V1, V1 | `etalcl + etalvc` block | 0.181, 0.134, 0.225 | Table 3 |
| omega^2 Ka | `etalka` | 0.0883 | Table 3 |
| omega^2 C0 | `etaleepo` | 0.501 | Table 3 |
| sigma_1^2 | `expSd` | 0.483 (SD = 0.695) | Table 3 |
| RBCPT | `lmtt` | 4.68 day | Table 4 |
| RBCLS | `lmtt_rbc` | 82.2 day | Table 4 |
| Smax | `lemax` | 0.437 | Table 4 |
| S50 | `lec50` | 3.68 ng/mL | Table 4 |
| Hb0 | `lrbase_hb` | 9.92 g/dL | Table 4 |
| theta_4 (PCNT on S50) | `e_conmed_platin_ec50` | 1.86 | Table 4 |
| omega^2 RBCLS | `etalmtt_rbc` | 1.63 | Table 4 |
| omega^2 S50, S50-Hb0, Hb0 | `etalec50 + etalrbase_hb` block | 8.05, 0.0539, 0.00720 | Table 4 |
| sigma_2^2 | `addSd_hb` | 0.401 (SD = 0.633 g/dL) | Table 4 |

# Part 1 -- Pharmacokinetics

## Replicating Figure 2C

Figure 2C of Agoram 2006 overlays the mean model prediction on observed mean
concentrations after a single 2.25 ug/kg IV dose, a single 2.25 ug/kg SC dose,
and a single 6.75 ug/kg SC dose, over 21 days. The panel is a typical-value
prediction, so between-subject variability is zeroed.

```{r figure-2c-events}
REF_WT <- 70                      # reference / typical weight, Agoram 2006 Eq 12-13

pk_arms <- tibble::tribble(
  ~arm,               ~dose_ug_kg, ~cmt,
  "2.25 ug/kg (IV)",  2.25,        "central",
  "2.25 ug/kg (SC)",  2.25,        "depot",
  "6.75 ug/kg (SC)",  6.75,        "depot"
)

# Dense early grid so Tmax and the alpha phase are resolved; 21 days matches
# the x-range of the published panel and keeps every concentration comfortably
# above solver round-off.
pk_times <- sort(unique(c(
  seq(0, 0.5, by = 0.005), seq(0.5, 3, by = 0.02),
  seq(3, 21, by = 0.1)
)))

make_pk_arm <- function(arm, dose_ug_kg, cmt, id) {
  dplyr::bind_rows(
    tibble(id = id, time = 0, amt = dose_ug_kg * REF_WT, evid = 1L, cmt = cmt),
    # Observation rows point at the ODE state `central`, never at the
    # algebraic observable `Cc`.
    tibble(id = id, time = pk_times, amt = NA_real_, evid = 0L, cmt = "central")
  ) |>
    dplyr::mutate(arm = arm, WT = REF_WT, CONMED_PLATIN_GT2 = 0)
}

pk_events <- dplyr::bind_rows(
  lapply(seq_len(nrow(pk_arms)), function(i) {
    make_pk_arm(pk_arms$arm[i], pk_arms$dose_ug_kg[i], pk_arms$cmt[i], id = i)
  })
)
stopifnot(!anyDuplicated(unique(pk_events[, c("id", "time", "evid")])))
```

```{r figure-2c-sim}
pk_typical <- rxode2::zeroRe(mod_pk)
sim_pk <- rxode2::rxSolve(
  pk_typical, events = pk_events, keep = c("arm", "WT"),
  omega = NA, sigma = NA, returnType = "data.frame"
)

# The ELISA cross-detects endogenous erythropoietin, so `Cc` is the TOTAL
# measured concentration (drug + eEPO). The drug-only profile is what the
# structural PK parameters describe, and it is what NCA must be run on -- an
# additive eEPO baseline that never decays would make AUCinf infinite.
sim_pk <- sim_pk |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(Cdrug = pmax(Cc - eepo, 0))
```

```{r figure-2c-plot}
ggplot(sim_pk, aes(time, Cc, colour = arm, linetype = arm)) +
  geom_line(linewidth = 0.7) +
  scale_y_log10(limits = c(0.1, 100)) +
  scale_x_continuous(breaks = seq(0, 22, by = 2)) +
  labs(
    x = "Time (days)", y = "Serum concentration (ng/mL)",
    colour = NULL, linetype = NULL,
    title = "Figure 2C -- typical-value profiles after a single dose",
    caption = "Replicates Figure 2C of Agoram 2006 (mean predicted profiles)."
  ) +
  theme_bw() + theme(legend.position = "top")
```

The published panel can be read off at a handful of landmarks. Digitised
values below are approximate (gridline resolution of a log-scale figure), so
they are checked with a tolerance rather than asserted exactly.

```{r figure-2c-landmarks}
landmark <- function(arm_name, t) {
  d <- sim_pk[sim_pk$arm == arm_name, ]
  d$Cc[which.min(abs(d$time - t))]
}

fig2c <- tibble::tibble(
  arm = c("2.25 ug/kg (IV)", "2.25 ug/kg (SC)", "6.75 ug/kg (SC)",
          "6.75 ug/kg (SC)", "6.75 ug/kg (SC)"),
  quantity = c("C at 5 min", "Cmax", "Cmax", "C at day 7", "C at day 21"),
  digitised = c(46, 6, 16, 7, 0.55),
  simulated = c(
    landmark("2.25 ug/kg (IV)", 5 / 1440),
    max(sim_pk$Cc[sim_pk$arm == "2.25 ug/kg (SC)"]),
    max(sim_pk$Cc[sim_pk$arm == "6.75 ug/kg (SC)"]),
    landmark("6.75 ug/kg (SC)", 7),
    landmark("6.75 ug/kg (SC)", 21)
  )
)

fig2c |>
  dplyr::mutate(pct_diff = 100 * (simulated - digitised) / digitised) |>
  dplyr::rename(
    "Arm" = arm, "Landmark" = quantity,
    "Digitised from Figure 2C (ng/mL)" = digitised,
    "Simulated (ng/mL)" = simulated,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = 2, caption = "Figure 2C landmarks, digitised vs simulated.")
```

```{r figure-2c-check}
chk <- 100 * abs(fig2c$simulated - fig2c$digitised) / fig2c$digitised
# Deterministic (typical-value) solve, so this bound does not depend on a
# simulated cohort. 25% is the resolution of reading a 3-decade log axis; a
# mis-transcribed volume, clearance or bioavailability moves these landmarks
# by a factor, not by a quarter.
stopifnot(max(chk) < 25)
```

## PKNCA validation

NCA is run on the baseline-corrected (drug-only) profile of the typical-value
solve. Running it over a full-IIV cohort would be misleading here: the
half-life estimate is `lambda_z`-selected per subject and returns `NA` for
subjects whose terminal phase is poorly resolved.

```{r pknca}
sim_nca <- sim_pk |>
  dplyr::filter(!is.na(Cdrug)) |>
  dplyr::select(id, time, arm, Cc = Cdrug)

# Guarantee a time = 0 record per (id, arm). For the SC arms the pre-dose
# drug concentration is genuinely 0; for the IV arm the t = 0 record already
# exists at the post-dose value and `distinct()` keeps it.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, arm, time)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(sim_nca), Cc ~ time | arm + id)

dose_df <- pk_events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, arm)
dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | arm + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE,
  half.life = TRUE, aucpext.obs = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(arm, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

nca_wide |>
  dplyr::select(arm, cmax, tmax, auclast, aucinf.obs, half.life, aucpext.obs) |>
  dplyr::rename(
    "Arm" = arm,
    "Cmax (ng/mL)" = cmax,
    "Tmax (day)" = tmax,
    "AUC0-21d (ng*day/mL)" = auclast,
    "AUC0-inf (ng*day/mL)" = aucinf.obs,
    "t1/2 (day)" = half.life,
    "AUC extrapolated (%)" = aucpext.obs
  ) |>
  knitr::kable(digits = 3, caption = "PKNCA summary of the typical-value single-dose profiles.")
```

Agoram 2006 reports no NCA table of its own -- it notes only that the model
parameters "generally agreed with previously reported noncompartmental
values". The NCA output is therefore checked against two identities that the
structural model must satisfy exactly. These are deterministic consequences of
a typical-value solve, not cohort statistics, so they are asserted tightly.

```{r pknca-identities}
get_nca <- function(arm_name, param) {
  v <- nca_wide[[param]][nca_wide$arm == arm_name]
  if (length(v) != 1L) stop("no unique NCA row for arm '", arm_name, "' / ", param)
  v
}

cl_typ    <- 2.010                              # L/day, Table 3, at WT = 70 kg, PCNT <= 2
dose_iv   <- 2.25 * REF_WT                      # ug
auc_iv    <- get_nca("2.25 ug/kg (IV)", "aucinf.obs")
auc_sc225 <- get_nca("2.25 ug/kg (SC)", "aucinf.obs")

identities <- tibble::tibble(
  Identity = c(
    "AUC0-inf (IV) = Dose / CL",
    "AUC0-inf (SC) / AUC0-inf (IV) = F"
  ),
  Expected = c(dose_iv / cl_typ, 0.443),
  Observed = c(auc_iv, auc_sc225 / auc_iv)
) |>
  dplyr::mutate(`Difference (%)` = 100 * (Observed - Expected) / Expected)

knitr::kable(identities, digits = 4, caption = "Closed-form identities the NCA output must satisfy.")

# Both sides come from the same deterministic solve, so the only error is
# numerical (trapezoidal AUC plus lambda_z extrapolation). 1.5% has ample
# headroom over that and still fails on a transcribed CL or F.
stopifnot(max(abs(identities$`Difference (%)`)) < 1.5)
# Terminal extrapolation must be small enough that AUC0-inf is meaningful.
stopifnot(max(nca_wide$aucpext.obs) < 1)
```

Both SC arms carry a terminal half-life longer than the IV arm
(`r sprintf("%.2f", get_nca("2.25 ug/kg (SC)", "half.life"))` vs
`r sprintf("%.2f", get_nca("2.25 ug/kg (IV)", "half.life"))` days), which is
the flip-flop behaviour Agoram 2006 attributes to absorption rate-limited
kinetics: `Ka` (0.318 /day) is smaller than the disposition `beta`, so the SC
terminal slope is the absorption rate constant.

# Part 2 -- Hemoglobin response

## Baseline is a true steady state

The paper writes the indirect-response baseline as `0 = Rin - Kout * Hb0`
(Equation 6) but implements the response as a transit chain, so `Rin` has to
be recovered from the chain's own drug-free steady state. With
`kPT = 1/RBCPT` and `kLS = 4/RBCLS`, that state is `A1 = Rin/kPT` and
`A2 = ... = A5 = Rin/kLS`, giving `Hb0 = 4*Rin/kLS = Rin*RBCLS`, hence
`Rin = Hb0/RBCLS`. If that derivation were wrong the model would drift away
from its own baseline with no drug on board -- which is exactly what this
check tests.

```{r ss-hold}
ss_events <- tibble(
  id = 1L, time = seq(0, 365, by = 7), amt = NA_real_, evid = 0L, cmt = "central",
  WT = REF_WT, CONMED_PLATIN = 0, CONMED_PLATIN_GT2 = 0
)

pd_typical <- rxode2::zeroRe(mod_pd)
ss <- rxode2::rxSolve(
  pd_typical, events = ss_events, omega = NA, sigma = NA, returnType = "data.frame"
) |>
  dplyr::filter(!is.na(hb))

stopifnot(nrow(ss) == length(unique(ss_events$time)))
ss_drift <- max(abs(ss$hb - 9.92))
cat(sprintf("Max drift from Hb0 = 9.92 g/dL over one year, drug-free: %.3g g/dL\n", ss_drift))
# Deterministic; the only error is solver tolerance.
stopifnot(ss_drift < 1e-6)
```

## Replicating Figure 4 -- external predictive check

Figure 4 shows the 2.5th-97.5th percentile band of predicted hemoglobin over
16 weeks in the 302-patient test set, whose "population characteristics and
administered doses [were] similar to those used in the modeling data set".
The regimens below are the three represented in the PD development set
(Agoram 2006 Table 1). The band is a *predictive* check, so it is taken from
the residual-error-carrying `sim` column rather than the individual
prediction.

```{r fig4-cohort}
# set.seed() seeds R's RNG, not rxode2's. rxode2 partitions its streams per
# solver thread, so this cohort is NOT reproducible across machines with
# different thread counts. Every assertion below is written to hold for any
# cohort the model can produce (see known-vignette-failure-patterns.md #12).
set.seed(20060901)
rxode2::rxSetSeed(20060901)

N_PER_ARM <- 150L                 # cap is 200 per arm
WEEKS     <- 16

pd_arms <- tibble::tribble(
  ~regimen,            ~dose_ug_kg, ~ii_days,
  "2.25 ug/kg QW",     2.25,        7,
  "4.5 ug/kg Q3W",     4.5,         21,
  "6.75 ug/kg Q3W",    6.75,        21
)

make_pd_arm <- function(regimen, dose_ug_kg, ii_days, id_offset) {
  # Body weight from the PD cohort: mean 69.9, SD 16.2, truncated to the
  # observed 39-136 kg range (Agoram 2006 Table 2).
  wt <- pmin(pmax(rnorm(N_PER_ARM, 69.9, 16.2), 39), 136)
  # 61% of the pooled cohort had a platinum-containing regimen; the paper does
  # not tabulate PCNT, so this split is an assumption (see Assumptions below).
  platin <- rbinom(N_PER_ARM, 1, 0.61)
  subj <- tibble(
    id = id_offset + seq_len(N_PER_ARM),
    WT = wt,
    CONMED_PLATIN = platin,
    # PCNT > 2 requires PCNT > 0; over a 16-week window a platinum patient on
    # a 3-weekly cycle receives more than two cycles, so the two indicators
    # coincide here. The mutual-consistency constraint is asserted below.
    CONMED_PLATIN_GT2 = platin,
    regimen = regimen
  )
  dose_times <- seq(0, WEEKS * 7 - 1, by = ii_days)
  doses <- subj |>
    tidyr::crossing(time = dose_times) |>
    dplyr::mutate(amt = dose_ug_kg * WT, evid = 1L, cmt = "depot")
  obs <- subj |>
    tidyr::crossing(time = seq(0, WEEKS * 7, by = 7)) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
  dplyr::bind_rows(doses, obs) |> dplyr::arrange(id, time, dplyr::desc(evid))
}

pd_events <- dplyr::bind_rows(
  make_pd_arm(pd_arms$regimen[1], pd_arms$dose_ug_kg[1], pd_arms$ii_days[1],   0L),
  make_pd_arm(pd_arms$regimen[2], pd_arms$dose_ug_kg[2], pd_arms$ii_days[2], 200L),
  make_pd_arm(pd_arms$regimen[3], pd_arms$dose_ug_kg[3], pd_arms$ii_days[3], 400L)
)
stopifnot(!anyDuplicated(unique(pd_events[, c("id", "time", "evid")])))
# Register constraint: CONMED_PLATIN_GT2 = 1 must imply CONMED_PLATIN = 1.
stopifnot(all(pd_events$CONMED_PLATIN_GT2 <= pd_events$CONMED_PLATIN))
```

```{r fig4-sim}
sim_pd <- rxode2::rxSolve(
  mod_pd, events = pd_events, keep = c("regimen", "WT", "CONMED_PLATIN"),
  returnType = "data.frame"
) |>
  dplyr::filter(!is.na(hb))

stopifnot(nrow(sim_pd) > 0, dplyr::n_distinct(sim_pd$id) == 3L * N_PER_ARM)
```

```{r figure-4-plot}
band <- sim_pd |>
  dplyr::mutate(week = time / 7) |>
  dplyr::group_by(week) |>
  dplyr::summarise(
    lo  = quantile(sim, 0.025, na.rm = TRUE),
    med = quantile(sim, 0.500, na.rm = TRUE),
    hi  = quantile(sim, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(band, aes(week, med)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "orange", alpha = 0.45) +
  geom_line(linewidth = 0.8) +
  scale_x_continuous(breaks = seq(0, 16, by = 5)) +
  coord_cartesian(ylim = c(5, 17)) +
  labs(
    x = "Time (weeks)", y = "Hemoglobin (g/dL)",
    title = "Figure 4 -- predicted hemoglobin, 2.5th-97.5th percentile band",
    caption = "Replicates Figure 4 of Agoram 2006 (predictive check band, pooled over regimens)."
  ) +
  theme_bw()
```

```{r figure-4-check}
wk <- function(w, col) band[[col]][which.min(abs(band$week - w))]

fig4 <- tibble::tibble(
  Quantity = c(
    "Median Hb at week 0 (g/dL)",
    "Median Hb at week 16 (g/dL)",
    "Rise in median Hb, week 0 to 16 (g/dL)",
    "Lower 2.5th percentile at week 16 (g/dL)",
    "Upper 97.5th percentile at week 16 (g/dL)"
  ),
  `Figure 4 (read off)` = c("~10", "~11.5-12", "~1.5-2", "~8.5", "~13.5-15"),
  Simulated = c(wk(0, "med"), wk(16, "med"), wk(16, "med") - wk(0, "med"),
                wk(16, "lo"), wk(16, "hi"))
)
knitr::kable(fig4, digits = 2, caption = "Figure 4 band, published vs simulated.")

# Cohort-derived, so these are magnitude bounds with headroom, not tight
# equalities. Week-0 median is anchored by Hb0 = 9.92 with a small IIV
# (omega^2 = 0.0072, ~8.5% CV) plus a 0.633 g/dL residual, so it is the
# stablest of the five.
#
# Realised at 1 / 2 / 4 / 8 solver threads (rxode2 partitions its RNG
# streams per thread, so a 2-core CI runner draws a different cohort than a
# workstation -- see known-vignette-failure-patterns.md #12):
#   med week 0  9.868 - 10.005
#   rise        1.187 -  1.250
#   med week 16 11.074 - 11.193
#   2.5th pct   8.473 -  8.709
#   97.5th pct  14.575 - 15.122
# Every bound below sits outside that range with at least ~0.5 g/dL of
# headroom, and each still goes red on a mis-transcribed Hb0, Smax, S50 or
# RBCLS, which move these numbers by whole g/dL. Do not tighten them back to
# whatever a single local run happens to give.
stopifnot(
  abs(wk(0, "med") - 9.92) < 0.5,
  wk(16, "med") - wk(0, "med") > 0.5,
  wk(16, "med") > 10.5, wk(16, "med") < 12.5,
  wk(16, "lo")  > 7.0,  wk(16, "lo")  < 10.0,
  wk(16, "hi")  > 12.0, wk(16, "hi")  < 17.0
)
```

The simulated band reproduces the published one in position and in shape: a
median that starts at the estimated `Hb0`, climbs over roughly the first 10
weeks and then plateaus, inside a band about 5-6 g/dL wide. The width is
driven almost entirely by `omega^2_S50 = 8.05` -- a CV above 5000% -- which
the paper attributes both to the genuine heterogeneity of the CIA population
and to `S50` absorbing the variability of `Smax`, whose IIV could not be
estimated.

## Time course of the response

```{r response-by-regimen}
sim_pd |>
  dplyr::mutate(week = time / 7) |>
  dplyr::group_by(regimen, week) |>
  dplyr::summarise(
    lo = quantile(hb, 0.25), med = quantile(hb, 0.50), hi = quantile(hb, 0.75),
    .groups = "drop"
  ) |>
  ggplot(aes(week, med, colour = regimen, fill = regimen)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.8) +
  labs(
    x = "Time (weeks)", y = "Hemoglobin (g/dL)", colour = NULL, fill = NULL,
    title = "Individual-prediction median and interquartile range by regimen"
  ) +
  theme_bw() + theme(legend.position = "top")
```

The lag before hemoglobin moves is set by `RBCPT` = 4.68 days (progenitor
maturation) followed by the mature-cell chain, whose transit time is
`RBCLS` = 82.2 days over four compartments. Agoram 2006 notes that 82.2 days
is well short of the ~120-day lifespan of healthy adult red cells, consistent
with the myelosuppressive effect of chemotherapy and anemia of chronic
disease.

## Assessment of covariate effects

Agoram 2006's "Assessment of Effect of Covariates" section predicts the
difference in 16-week hemoglobin response, at a fixed 150 ug QW dose
(equivalent to the prescribed 2.25 ug/kg QW for a 70 kg patient), between the
extremes of each covariate's range. The published mean (95% CI) is
0.939 (0.283, 1.45) g/dL for body weight and 0.317 (-0.504, 0.877) g/dL for
platinum-containing chemotherapy.

```{r covariate-effects}
cov_events <- function(wt, platin) {
  tibble(
    id = 1L, time = seq(0, 15 * 7, by = 7), amt = 150, evid = 1L, cmt = "depot",
    WT = wt, CONMED_PLATIN = platin, CONMED_PLATIN_GT2 = platin
  ) |>
    dplyr::bind_rows(
      tibble(
        id = 1L, time = seq(0, 16 * 7, by = 7), amt = NA_real_, evid = 0L,
        cmt = "central", WT = wt, CONMED_PLATIN = platin,
        CONMED_PLATIN_GT2 = platin
      )
    ) |>
    dplyr::arrange(time, dplyr::desc(evid))
}

delta_hb <- function(wt, platin) {
  s <- rxode2::rxSolve(
    pd_typical, events = cov_events(wt, platin),
    omega = NA, sigma = NA, returnType = "data.frame"
  ) |>
    dplyr::filter(!is.na(hb))
  s$hb[which.min(abs(s$time - 112))] - s$hb[which.min(abs(s$time - 0))]
}

eff_wt     <- delta_hb(39, 0) - delta_hb(136, 0)
eff_platin <- delta_hb(REF_WT, 0) - delta_hb(REF_WT, 1)

cov_tab <- tibble::tibble(
  Covariate = c("Body weight, 39 kg vs 136 kg", "Platinum chemotherapy, no vs yes"),
  `Published mean (g/dL)` = c(0.939, 0.317),
  `Published 95% CI` = c("(0.283, 1.45)", "(-0.504, 0.877)"),
  `Typical-value contrast (g/dL)` = c(eff_wt, eff_platin),
  `Inside published CI` = c(
    eff_wt > 0.283 && eff_wt < 1.45,
    eff_platin > -0.504 && eff_platin < 0.877
  )
)
knitr::kable(cov_tab, digits = 3, caption = "16-week hemoglobin response difference at 150 ug QW, at the extremes of each covariate.")

# Deterministic typical-value contrasts, so this gate is stable. The paper's
# figures are Monte Carlo means over the full-IIV cohort rather than
# typical-value contrasts, which is why the point estimates differ while both
# land inside the published intervals.
stopifnot(all(cov_tab$`Inside published CI`))
```

Both effects reproduce inside the paper's own published confidence intervals.
The paper's conclusion follows directly from the width of those intervals:
"it was not possible to classify any of the covariates as clinically
significant", and dose adjustment on either covariate is not recommended.

# Assumptions and deviations

- **Residual-error form for hemoglobin.** Table 4 reports `sigma_2^2 = 0.401`
  without naming an error model, and Equation 2's log-transformed exponential
  form is stated only for the *pharmacokinetic* observations. It is encoded
  here as **additive**, SD = `sqrt(0.401)` = 0.633 g/dL, which is about 6% of
  the 9.92 g/dL baseline and matches the roughly +/- 1 g/dL scatter about the
  line of identity in Figure 3A. A log-scale reading of the same variance
  would be a 63% CV and would spread individual predictions over roughly
  3-34 g/dL, which Figure 3A rules out.
- **Equations 7 and 8 as printed are mutually inconsistent.** Equation 8
  defines `Stim = (1 + Smax*C_DA/(S50 + C_DA))` -- the `1 +` is inside the
  parentheses -- while Equation 7 is typeset as `dA1/dt = Rin(1 = Stim) - kPT*A1`,
  where `1 = Stim` is plainly a typographical error. Under either repair
  (`Rin * Stim` with the `1 +` inside `Stim`, or `Rin * (1 + Stim)` with
  `Stim = Smax*C_DA/(S50+C_DA)`) the production term is identical:
  `Rin * (1 + Smax*C_DA/(S50 + C_DA))`, saturating at `1 + Smax` = 1.437. That
  is the form encoded, and it is the one the abstract describes as an
  "incremental maximum stimulation of hemoglobin production [of] 43.7%".
  Equation 10 is likewise printed with a left-hand side of `dA2/dt` where the
  index range `i = 3,4,5` on the right makes clear it is `dAi/dt`.
- **`Rin` is derived, not published.** The paper does not print a value for
  `Rin` or `Kout`; only `Hb0` and `RBCLS` are estimated. `Rin = Hb0 / RBCLS`
  is the drug-free steady state of the published chain, verified above to hold
  the baseline to within solver tolerance over a simulated year.
- **Intercompartmental clearance `Q` is encoded as `fixed()`.** Its Table 3
  row is the only one with both a blank standard-error column and a blank
  bootstrap column; every parameter the paper estimated has both. The paper
  does not say in prose that `Q` was fixed, so this is an inference from the
  table's structure. It has no effect on any simulation -- `fixed()` is
  provenance metadata only -- but a reviewer with access to the control stream
  should confirm it.
- **`V2` and `Q` carry no IIV, and `F` carries none either.** This follows the
  paper: "It was not possible to estimate the IIV of Q and V2", and IIV of `F`
  was not estimated because the IV and SC data came from separate patients.
- **No IIV on `RBCPT` or `Smax`.** The paper states both explicitly. The
  consequence, which it also flags, is that the estimated `S50` variability
  probably absorbs `Smax` variability, so covariate conclusions "may equally
  apply to both these parameters".
- **The abstract's "steady-state volume of distribution, 3390 mL" is the
  central volume `V1`, not `Vss`.** Table 3 labels 3390 mL as `theta_V1`;
  `Vss = V1 + V2` = 3641 mL. The table is taken as authoritative.
- **PD covariate distributions in the virtual cohort are assumptions.** Body
  weight is drawn as N(69.9, 16.2) truncated to the published 39-136 kg range
  (Table 2). The paper never tabulates `PCNT`, so the proportion of patients on
  platinum-containing chemotherapy is set to 61%, chosen so that the
  non-platinum tumor types the paper names as typically platinum-free (breast,
  and the pancreatic / prostate cases it groups under "other") are the
  complement. This affects only the width and position of the Figure 4 band,
  not any parameter.
- **The `CONMED_PLATIN` / `CONMED_PLATIN_GT2` pair.** Agoram 2006 dichotomises
  a single underlying count, `PCNT`, at two different thresholds in the same
  publication: `> 2` for the PK model's clearance effect (Equation 12) and
  `> 0` for the PD model's `S50` effect (Equation 14). Per the standing
  count-covariate policy the count is carried as two binary indicator columns
  rather than as an integer, and a data set supporting both models must
  provide both. In the virtual cohort above the two coincide, because a
  patient on a 3-weekly platinum regimen receives more than two cycles inside
  a 16-week window.
- **Figure 2C landmarks are digitised.** The published panel reports no
  numeric table, so the five landmark values in the comparison table were read
  off the log-scale figure and are accurate only to gridline resolution; they
  are checked with a 25% tolerance.
- **No published NCA table exists for this paper.** The PK section therefore
  validates the NCA output against closed-form identities the structural model
  must satisfy (`AUC0-inf = Dose/CL` for the IV arm, and the SC/IV AUC ratio
  equalling `F`) rather than against transcribed reference values.
