---
title: "Cyclophosphamide (Ahmed 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cyclophosphamide (Ahmed 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Ahmed_2020_cyclophosphamide"))$reference`
* Description: `r rxode2::rxode(readModelDb("Ahmed_2020_cyclophosphamide"))$description`
* Article: <https://doi.org/10.3389/fphar.2020.00406> (open access)

A literature check on 2026-09-26 (Europe PMC) found no erratum or correction
for this article. The article's supplementary files are the figure images only;
there is no NONMEM control stream.

# Population

Ahmed 2020 enrolled 267 women with breast cancer at the radiotherapy centre of
Tikur Anbessa Specialized Hospital, Addis Ababa, Ethiopia, who were starting
their first cycle of cyclophosphamide-containing chemotherapy. Median age was 38
years (IQR 33-48), mean body surface area (BSA) 1.59 +/- 0.20 m^2 and mean body
mass index 23.78 +/- 4.8 kg/m^2 (Table 1). Patients received cyclophosphamide as
a 30-min IV infusion either at 600 mg/m^2 (AC or AC-T regimens; 161 patients,
60.3%) or at 500 mg/m^2 (FAC regimen, which adds 5-fluorouracil; 106 patients,
39.7%). The median absolute dose was 930 mg (650-1,150 mg) in the 600 mg/m^2
group and 777.5 mg (600-1,000 mg) in the 500 mg/m^2 group. The analysis used
532 plasma concentrations (about two per patient, mostly at 1-6 h after the
start of infusion; 17 patients were sampled up to 22 h) and absolute neutrophil
counts (ANC) at baseline and on day 20. Median baseline ANC was 3645.5
cells/mm^3 (Table 1; the table header says 10^3 cells/mm^3 but the values are
cells/mm^3).

The same information is available programmatically via
`readModelDb("Ahmed_2020_cyclophosphamide")()$population`.

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Structure: one compartment, IV infusion, first-order elimination | -- | Results 'Population Pharmacokinetic Modeling' |
| `lcl` (CL, 600 mg/m^2 regimen) | 5.41 L/h | Table 3, final model |
| `lvc` (VD, 600 mg/m^2 regimen, BSA 1.58 m^2) | 46.5 L | Table 3, final model |
| `e_dose_high_cl` (500 mg/m^2 regimen on CL, `1 + THETA`) | -0.323 | Table 4 |
| `e_dose_high_vc` (500 mg/m^2 regimen on VD, `1 + THETA`) | -0.371 | Table 4 |
| `e_bsa_vc` (`(BSA/1.58)^THETA` on VD) | 0.861 | Table 4; Methods Equation 1 |
| `etalcl` | 46.4% CV -> 0.19500 | Table 3, final model; `%CV = sqrt(exp(omega^2) - 1)` (Methods) |
| `etalvc` | 35.9% CV -> 0.12123 | Table 3, final model |
| `addSd` | 1.54 mg/L | Table 3, 'Additive error 1' |
| PD structure: `ANC = ANC0 + slope * AUC` | -- | Methods Equation 5; Results 'Pharmacodynamic Modeling'; Figure 7 |
| `lrbase_anc` (ANC0) | 3450 cells/mm^3 | Results 'Pharmacodynamic Modeling' |
| `slope_anc` | -1.42 (cells/mm^3)/(umol*h/L) | Results 'Pharmacodynamic Modeling' |
| `propSd_ANC`, `addSd_ANC` | 0 (fixed) | Combined error selected, magnitudes not reported |
| AUC unit conversion, MW 261.09 g/mol | -- | Table 3 reports AUC in umol*h/L; molecular weight of anhydrous cyclophosphamide |

# Virtual cohort

Two arms of 200 patients each, one per regimen. BSA is drawn from a normal
distribution with the Table 1 mean and SD, truncated to 1.2-2.2 m^2; the
absolute dose is the regimen dose per m^2 times the patient's BSA, infused over
30 min.

```{r cohort}
set.seed(20200423)
rxode2::rxSetSeed(20200423)
n_per_arm <- 200

cohort <- tibble::tibble(
  id = seq_len(2 * n_per_arm),
  arm = rep(c("500 mg/m^2", "600 mg/m^2"), each = n_per_arm),
  DOSE_HIGH = rep(c(0L, 1L), each = n_per_arm),
  BSA = pmin(pmax(rnorm(2 * n_per_arm, 1.59, 0.20), 1.2), 2.2)
) |>
  dplyr::mutate(dose = ifelse(DOSE_HIGH == 1L, 600, 500) * BSA)

obs_times <- c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 5, 6, 8, 10, 12, 16, 22, 24, 36, 48, 480)

make_events <- function(cohort, obs_times) {
  doses <- cohort |>
    dplyr::transmute(
      id, time = 0, amt = dose, rate = dose / 0.5, evid = 1L,
      cmt = "central", dvid = NA_integer_, DOSE_HIGH, BSA
    )
  obs <- tidyr::expand_grid(id = cohort$id, time = obs_times) |>
    dplyr::left_join(cohort |> dplyr::select(id, DOSE_HIGH, BSA), by = "id") |>
    dplyr::mutate(
      amt = NA_real_, rate = NA_real_, evid = 0L,
      cmt = NA_character_, dvid = 1L
    )
  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    as.data.frame()
}

events <- make_events(cohort, obs_times)
```

# Simulation

```{r simulate}
mod <- readModelDb("Ahmed_2020_cyclophosphamide")

sim <- rxode2::rxSolve(
  mod, events,
  keep = c("DOSE_HIGH", "BSA"),
  returnType = "data.frame", useLinCmt = FALSE
) |>
  dplyr::left_join(cohort |> dplyr::select(id, arm), by = "id")

# Typical-value solve (no between-subject variability, no residual error)
sim_tv <- rxode2::rxSolve(
  mod, events,
  keep = c("DOSE_HIGH", "BSA"),
  returnType = "data.frame", useLinCmt = FALSE,
  omega = NA, sigma = NA
) |>
  dplyr::left_join(cohort |> dplyr::select(id, arm), by = "id")
```

## Typical-value check against the closed form

A one-compartment infusion model has an analytic solution. At the end of a
0.5 h infusion of dose `D` the concentration is
`D / (0.5 * CL) * (1 - exp(-k * 0.5))` with `k = CL / V`, and the day-20 AUC is
`D / CL`. Both sides use the same parameters, so the difference is numerical
error only and a tight bound is appropriate.

```{r closed_form}
tv_par <- cohort |>
  dplyr::mutate(
    cl = 5.41 * (1 - 0.323 * (1 - DOSE_HIGH)),
    vc = 46.5 * (BSA / 1.58)^0.861 * (1 - 0.371 * (1 - DOSE_HIGH)),
    k = cl / vc,
    c_eoi = dose / (0.5 * cl) * (1 - exp(-k * 0.5)),
    auc_umol = dose / cl * 1000 / 261.09,
    anc_d20 = 3450 - 1.42 * auc_umol
  )

chk <- sim_tv |>
  dplyr::filter(time %in% c(0.5, 480)) |>
  dplyr::select(id, time, Cc, ANC) |>
  tidyr::pivot_wider(names_from = time, values_from = c(Cc, ANC)) |>
  dplyr::left_join(tv_par, by = "id")

stopifnot(
  all(abs(chk$Cc_0.5 / chk$c_eoi - 1) < 1e-4),
  all(abs(chk$ANC_480 - chk$anc_d20) < 0.5),
  all(abs(sim_tv$ANC[sim_tv$time == 0] - 3450) < 1e-8)
)

tv_par |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    `CL (L/h)` = signif(unique(cl), 3),
    `median V (L)` = signif(median(vc), 3),
    `median t1/2 (h)` = signif(median(log(2) / k), 3),
    .groups = "drop"
  ) |>
  dplyr::rename(Regimen = arm) |>
  knitr::kable(caption = "Typical-value PK parameters by regimen in the virtual cohort.")
```

# Replicate published figures

The paper has no concentration-time profile figure beyond the pcVPC of
Figure 3, whose axes are prediction corrected. The panel below shows the
simulated 5th, 50th and 95th percentiles over the sampling window used in the
study (up to 22 h after the start of infusion).

```{r vpc, fig.cap = "Simulated cyclophosphamide concentrations (median and 90% interval) over the Ahmed 2020 sampling window, by regimen; compare the range of Figure 3 of Ahmed 2020."}
sim |>
  dplyr::filter(time <= 24) |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(
    lo = quantile(Cc, 0.05), med = median(Cc), hi = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, med, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, colour = NA) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Time after start of infusion (h)", y = "Cyclophosphamide (mg/L)",
       colour = "Regimen", fill = "Regimen")
```

```{r anc_fig, fig.cap = "Model-predicted day-20 ANC against the cyclophosphamide AUC; replicates the range of the population predictions in Figure 7 of Ahmed 2020 (baseline predictions at 3450 cells/mm^3, day-20 predictions mostly between about 2000 and 3100 cells/mm^3)."}
anc_d20 <- sim |>
  dplyr::filter(time == 480) |>
  dplyr::mutate(auc_umol = auc * 1000 / 261.09)

ggplot(anc_d20, aes(auc_umol, ANC, colour = arm)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 3450, linetype = 2) +
  labs(x = "Cyclophosphamide AUC0-day20 (umol*h/L)",
       y = "Predicted ANC on day 20 (cells/mm^3)", colour = "Regimen")
```

# PKNCA validation

Concentrations are converted to umol/L so that the AUC is on the umol*h/L
scale Ahmed 2020 Table 3 uses. The paper's derived PK summaries (Table 3) are
for the whole cohort, so the NCA is also run on the pooled cohort, grouped as
`All`, next to the per-regimen results.

```{r pknca}
# Late samples (36 h onward) sit at the solver's numerical noise floor and can
# be negative by ~1e-30; PKNCA returns NaN for AUClast on any profile with a
# negative concentration, so those are floored at zero.
conc_df <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(conc_umol = pmax(Cc, 0) * 1000 / 261.09) |>
  dplyr::select(id, time, conc_umol, arm)

dose_df <- cohort |>
  dplyr::transmute(id, time = 0, amt = dose, arm)

run_nca <- function(conc_df, dose_df) {
  conc_obj <- PKNCA::PKNCAconc(conc_df, conc_umol ~ time | arm + id)
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id,
                               route = "intravascular", duration = 0.5)
  intervals <- data.frame(
    start = 0, end = 480,
    cmax = TRUE, tmax = TRUE, auclast = TRUE, half.life = TRUE
  )
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}

nca_arm <- run_nca(conc_df, dose_df)
nca_all <- run_nca(conc_df |> dplyr::mutate(arm = "All"),
                   dose_df |> dplyr::mutate(arm = "All"))

summary(nca_arm)
summary(nca_all)
```

# Comparison against published values

Ahmed 2020 Table 3 reports the median of the empirical Bayes AUC0-day20
(565.7 umol*h/L) and half-life (5.48 h) for the whole cohort.

```{r nca_compare}
published <- tibble::tribble(
  ~arm, ~auclast, ~half.life,
  "All", 565.7, 5.48
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_all,
  reference = published,
  by = "arm",
  units = c(auclast = "umol*h/L", half.life = "h"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Simulated vs. published (Ahmed 2020 Table 3). * differs from reference by >20%.")
```

The simulated median half-life agrees with Table 3. The simulated median AUC is
about 30% higher than the 565.7 umol*h/L of Table 3. AUC0-day20 is `dose / CL`
independent of the volume, and the typical-value AUC alone is about 660
umol*h/L in the 600 mg/m^2 arm (930 mg / 5.41 L/h) and about 810 umol*h/L in
the 500 mg/m^2 arm, so the gap is a property of the published parameter set
and not of the simulation. Table 3 does not state how its derived AUC was
computed; it was not used to adjust any parameter.

The regimen-specific clearances reported in the Results (mean 3.97 L/h for the
500 mg/m^2 regimen and 5.81 L/h for the 600 mg/m^2 regimen) and the
600 mg/m^2-regimen half-lives by BSA group (4.78 h below 1.5 m^2, 5.42 h for
1.5-1.74 m^2, 6.71 h above 1.75 m^2) are empirical Bayes summaries; the
simulated equivalents are shown below. With about two samples per patient the
empirical Bayes estimates are shrunk towards the typical value, so the
simulated spread across BSA groups is expected to be at least as wide as the
published one; the published half-lives are not stated as means or medians.

```{r ebe_compare}
ind <- sim |>
  dplyr::filter(time == 0.5) |>
  dplyr::select(id, arm, BSA, cl, vc, kel)

ind |>
  dplyr::group_by(arm) |>
  dplyr::summarise(`mean CL (L/h)` = signif(mean(cl), 3), .groups = "drop") |>
  dplyr::mutate(`published mean CL (L/h)` = c(3.97, 5.81)) |>
  dplyr::rename(Regimen = arm) |>
  knitr::kable(caption = "Mean individual clearance by regimen (Ahmed 2020 Results).")

ind |>
  dplyr::filter(arm == "600 mg/m^2") |>
  dplyr::mutate(bsa_group = cut(BSA, c(0, 1.5, 1.75, Inf),
                                labels = c("< 1.5", "1.5-1.74", ">= 1.75"),
                                right = FALSE)) |>
  dplyr::group_by(bsa_group) |>
  dplyr::summarise(`simulated median t1/2 (h)` = signif(median(log(2) / kel), 3),
                   .groups = "drop") |>
  dplyr::mutate(`published t1/2 (h)` = c(4.78, 5.42, 6.71)) |>
  dplyr::rename(`BSA group (m^2)` = bsa_group) |>
  knitr::kable(caption = "Half-life by BSA group in the 600 mg/m^2 regimen (Ahmed 2020 Results).")

stopifnot(
  # Structural: the regimen effect on CL reproduces the reported ratio of
  # mean clearances (3.97 / 5.81 = 0.68 vs 1 - 0.323 = 0.677).
  abs(median(ind$cl[ind$arm == "500 mg/m^2"]) /
        median(ind$cl[ind$arm == "600 mg/m^2"]) / 0.677 - 1) < 0.2,
  abs(median(log(2) / ind$kel) / 5.48 - 1) < 0.2
)
```

# Assumptions and deviations

* **Sign of the PD slope.** Equation 5 is printed as `Y = ANCo - slope * AUC`
  and the Results give `slope = -1.42`. Taken literally the count would rise
  with exposure, which contradicts the Results ("a one unit increase in the AUC
  of CPA was associated with a decrease in the neutrophil count by 1.42"), the
  Discussion, and Figure 7, where every population prediction lies at or below
  3450. The model therefore uses `ANC = ANC0 + slope * AUC` with the reported
  `slope = -1.42`.
* **AUC unit in the PD model.** The paper does not state the AUC unit in the
  PD fit. Table 3 reports AUC0-day20 in umol*h/L, and only that unit reproduces
  the Figure 7 population predictions (about 2000-3100 cells/mm^3 on day 20
  for an interquartile AUC of 445-828 umol*h/L). The concentration state is in
  mg/L, so the AUC is converted with the cyclophosphamide molecular weight
  261.09 g/mol, which the paper does not print.
* **AUC regressor over time.** The model regresses the ANC on the running AUC
  integral, so it returns ANC0 at baseline and the Equation 5 value on day 20
  (by which time the AUC is complete). The paper only fitted those two time
  points; predictions at intermediate times have no source support.
* **PD residual error and variability.** A combined proportional and additive
  residual error was selected for ANC but its magnitudes are not reported, so
  both are fixed to 0. No between-subject variability on ANC0 or the slope is
  reported; Figure 7 shows individual and population predictions that coincide,
  consistent with none being estimated.
* **Second additive PK residual.** Table 3 lists a second additive error of
  0.0001 mg/mL (0.1 mg/L) whose role is not described. It is not encoded; added
  in quadrature to 1.54 mg/L it would change the residual SD by 0.2%.
* **Residual magnitude as an SD.** The 1.54 mg/L is taken as a standard
  deviation because it is reported in concentration units.
* **BSA reference.** Methods Equation 1 describes the BSA normalising value as
  the median; Table 4 prints 1.58 m^2, which is used.
* **Regimen covariate.** The 500 mg/m^2 (FAC) and 600 mg/m^2 (AC, AC-T)
  regimens differ in co-administered drugs as well as in cyclophosphamide dose,
  so the covariate is a regimen effect. It is carried in the `DOSE_HIGH`
  column (1 = 600 mg/m^2), with the Table 4 coefficient applied to the
  500 mg/m^2 complement.
* **Genotype effects.** CYP3A5 and CYP2C9 genotype associations are post hoc
  analyses of empirical Bayes estimates and are not part of the NONMEM model;
  they are documented in `covariatesDataExcluded`.
* **Virtual cohort.** BSA is drawn from a truncated normal with the Table 1
  mean and SD; the paper gives no BSA distribution by regimen.
