---
title: "Reserpine rat PK-PD (AlfoseaCuadrado 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reserpine rat PK-PD (AlfoseaCuadrado 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
mod <- readModelDb("AlfoseaCuadrado_2024_reserpine_rat")
ui <- rxode2::rxode2(mod)
```

* Citation: `r ui$reference`
* Description: `r ui$description`
* Article: <https://doi.org/10.3390/pharmaceutics16081101>
* Analysis dataset (deposited by the authors): <https://doi.org/10.5281/zenodo.11206173>

Reserpine blocks the vesicular monoamine transporter 2 (VMAT-2), depleting
serotonin, dopamine and noradrenaline. Repeated low-dose reserpine in the rat is
the reserpine-induced myalgia (RIM) model, the most widely used animal model of
fibromyalgia syndrome. This paper is the first population PK-PD description of
that model.

# Population

120 male Sprague-Dawley rats weighing 300-450 g received reserpine 0.1, 0.5 or
1 mg/kg once daily for three consecutive days (Section 2.1). Sampling was
**destructive** -- one terminal sample per animal, taken pre-dose or 30 min, 2,
4, 24 or 48 h after the *third* dose, i.e. 48-96 h after the first (Section 2.1,
Figure S1). 120 plasma reserpine and 828 monoamine observations were collected;
Table 1 gives the counts per dose group, neurotransmitter and tissue.

Monoamines were measured in three nervous-tissue regions: amygdala (AMY), medial
prefrontal cortex (PFC) and lumbar spinal cord (SC). Because each animal
contributes a single time point, the authors assembled longitudinal profiles by
pooling samples across animals (Section 3.1), and identify this as the reason
for the unusually large inter-animal variability (Section 4).

The same information is available programmatically via `ui$meta$population`.

```{r population}
str(ui$meta$population, max.level = 1)
```

# Structural reconstruction

**The paper prints no model equations.** Section 3.2.2 describes the structure in
prose, Figure 1 gives a schematic, and Table 2 lists the parameter estimates. The
ODE system in the packaged model was therefore reconstructed from Figure 1 plus
the Table 2 footnote, and then *falsified* against the paper's own quantitative
output (Figure 3 profiles and the Figure 4 / Section 3.3 numerical predictive
check). The checks in this vignette are that falsification.

The reconstructed system is the precursor-pool model of Sharma et al. 1998
(the paper's reference 40) with a parallel three-transit chain:

$$\frac{dP}{dt} = k_{in} - k_p\,(1 + SLP_1 \cdot C_c)\,P, \qquad P(0) = k_{in}/k_p$$
$$\frac{dR}{dt} = k_p\,(1 + SLP_1 \cdot C_c)\,P - k_{out}\,M_3\,R, \qquad R(0) = k_{in}/k_{out}$$
$$\frac{dM_1}{dt} = k_0\,(1 + SLP_2 \cdot C_c) - k_{tr}\,M_1, \qquad M_1(0)=M_2(0)=M_3(0)=1$$

with $M_2$, $M_3$ following the chain. Reserpine ($C_c$) stimulates the
precursor-to-response transfer through $SLP_1$ (the green "+" on $k_p$ in
Figure 1) and, with a delay imposed by the transit chain, the degradation of the
response through $SLP_2$ acting on $k_0$ (the green "+" on $k_0$). Setting
$k_{tr} = k_0$ is what makes the paper's stated initial condition
$M_1(0)=M_2(0)=M_3(0)=1$ a steady state, which is why Table 2 reports a single
value for the chain.

In package naming these map onto the registered precursor-pool canonicals: the
paper's $k_{in}$ (zero-order production of the precursor) is `kpin`, its $k_p$
(precursor to response) is `kin`, and its $k_{out}$ is `kout` -- the same
convention as the library template `indirect_prec_1cpt_stim_CLV`.

# Source trace

| Equation / parameter | Value | Source location |
|---|---|---|
| `lka` (ka1) | 19.14 1/h, fixed | Table 2, row `ka1`, marked FIX |
| `ld1` (Tk0_2) | 45.43 h | Table 2, row `ka2`; defined as a *duration* by the Table 2 footnote and the Figure 1 legend |
| `logitfdepot` (F1) | 0.95 | Table 2, row `F1` |
| `lvc` (V) | 1.3 mL/kg = 1.3e-3 L/kg | Table 2, row `V` (unit conversion) |
| `lcl` (CL) | 4.5e-1 mL/h/kg = 4.5e-4 L/h/kg | Table 2, row `CL` (unit conversion) |
| `lkpin` (kin, AMY) | 6.97 | Table 2, row `kin (mg/h) AMY` |
| `lkpin_pfc` (kin, PFC) | 2.10 | Table 2, row `kin (mg/h) PFC` |
| `lkpin_sc` (kin, SC) | 1.78 | Table 2, row `kin (mg/h) SC` |
| `lkin` (kp) | 8.6e-4 1/h | Table 2, row `kp` |
| `lkout` (kout) | 2.7e-2 1/h | Table 2, row `kout` |
| `lktr` (k0) | 1.9e-1 1/h | Table 2, row `k0` |
| `lslp1` (SLP1) | 1.1e-1 | Table 2, row `SLP1` |
| `lslp2` (SLP2) | 1.25 | Table 2, row `SLP2` |
| IIV (all `eta*`) | CV% -> `omega^2 = log(1 + CV^2)` | Table 2, "Inter-Animal Variability" column |
| `propSd` (PK RUV) | 0.54 | Table 2, "Residual unexplained variability", PK (%) |
| `propSd_effect` (PD RUV) | 0.71 | Table 2, "Residual unexplained variability", PD (%) |
| PK structure (two parallel depots) | n/a | Figure 1, left panel; Section 3.2.1 |
| PD structure (precursor pool + transit chain) | n/a | Figure 1, right panel; Section 3.2.2 |
| Region covariate on kin | n/a | Section 3.2.2, "brain regions ... as statistically significant covariates on kin" |

# Simulation setup

The model declares two endpoints (`Cc` and `effect`). In rxode2 5.1.3 the
modelled `dur(central)` is not evaluated for a dose record flagged `rate = -2`
in that configuration, so the zero-order absorption duration is supplied on the
dose record instead; the two are numerically identical.

```{r sim-helpers}
TK02 <- 45.43   # Table 2 zero-order absorption duration (h)

# Three daily doses; observations are placed on the `effect` ODE state, which is
# a declared endpoint. rxode2 returns the algebraic observable Cc on those rows
# too, so both endpoints are available from one grid.
make_events <- function(dose, times, n = 1L) {
  ev <- rxode2::et(amt = dose, time = c(0, 24, 48), cmt = "depot")
  ev <- rxode2::et(ev, amt = dose, time = c(0, 24, 48), cmt = "central", dur = TK02)
  ev <- rxode2::et(ev, times, cmt = "effect")
  if (n > 1L) ev <- rxode2::et(ev, id = seq_len(n))
  ev
}

REGION <- list(
  AMY = c(CNSREG_PFC = 0, CNSREG_SC = 0),
  PFC = c(CNSREG_PFC = 1, CNSREG_SC = 0),
  SC  = c(CNSREG_PFC = 0, CNSREG_SC = 1)
)

# The stiff post-dose transient (SLP2 * Cc reaches ~10^3) needs a high-order
# solver; the default settings drop a minority of animals in the population run.
SOLVER <- list(method = "dop853", maxsteps = 5e6L)

solve_typical <- function(dose, region, times) {
  do.call(
    rxode2::rxSolve,
    c(list(rxode2::zeroRe(ui), events = make_events(dose, times),
           params = REGION[[region]], returnType = "data.frame"), SOLVER)
  )
}
```

# Baselines are the algebraic steady state

Before any drug is given the system must sit exactly at its steady state. With
$P(0)=k_{in}/k_p$ and $R(0)=k_{in}/k_{out}$ this is an exact identity, so it is
asserted rather than eyeballed.

```{r baselines}
kpin <- c(AMY = 6.97, PFC = 2.10, SC = 1.78)   # Table 2
kp   <- 8.6e-4
kout <- 2.7e-2

baseline <- lapply(names(REGION), function(rg) {
  s <- solve_typical(1, rg, c(0, 1))
  data.frame(
    Region        = rg,
    precursor1_0  = s$precursor1[s$time == 0],
    expected_P0   = kpin[[rg]] / kp,
    effect_0      = s$effect[s$time == 0],
    expected_R0   = kpin[[rg]] / kout
  )
}) |> dplyr::bind_rows()

stopifnot(
  all(abs(baseline$precursor1_0 / baseline$expected_P0 - 1) < 1e-6),
  all(abs(baseline$effect_0     / baseline$expected_R0 - 1) < 1e-6)
)

baseline |>
  dplyr::rename(
    "Precursor pool P(0)"   = precursor1_0,
    "Expected k_in/k_p"     = expected_P0,
    "Response R(0)"         = effect_0,
    "Expected k_in/k_out"   = expected_R0
  ) |>
  knitr::kable(digits = 2,
               caption = "Simulated baselines equal the algebraic steady state exactly.")
```

The amygdala precursor pool starts at `r round(kpin[["AMY"]] / kp)`, matching the
~8000 read off the top-left panel of Figure 3 and the paper's observation
(Section 4) that "the more remarkable MA synthesis of the analyzed centers is
found in the AMY precursor pool".

# Replicating Figure 3

Figure 3 shows typical-value profiles for the three studied doses (0.1, 0.5,
1 mg/kg, solid) plus two untested doses (2 and 4 mg/kg, dotted), in each region.

```{r figure-3, fig.height = 6}
grid_fine <- seq(0, 96, by = 0.5)
doses <- c(0.1, 0.5, 1, 2, 4)

fig3 <- lapply(names(REGION), function(rg) {
  lapply(doses, function(d) {
    s <- solve_typical(d, rg, grid_fine)
    data.frame(Region = rg, Dose = d, time = s$time,
               Response = s$effect, Precursor = s$precursor1)
  }) |> dplyr::bind_rows()
}) |> dplyr::bind_rows()

fig3 |>
  tidyr::pivot_longer(c(Response, Precursor), names_to = "State", values_to = "value") |>
  dplyr::mutate(Region = factor(Region, c("AMY", "PFC", "SC")),
                Dose = factor(Dose)) |>
  ggplot(aes(time, value, colour = Dose, linetype = Dose %in% c("2", "4"))) +
  geom_line(linewidth = 0.6) +
  facet_grid(State ~ Region, scales = "free_y") +
  scale_linetype_manual(values = c("solid", "22"), guide = "none") +
  scale_colour_viridis_d(end = 0.9, direction = -1) +
  labs(x = "Time (h)", y = "Monoamine level (mg/L)",
       caption = "Replicates Figure 3 of Alfosea-Cuadrado 2024. Dotted = untested doses.")
```

Selected values against the figure, for the amygdala:

```{r figure-3-values}
fig3 |>
  dplyr::filter(Region == "AMY", time == 96) |>
  dplyr::transmute(`Dose (mg/kg)` = Dose,
                   `Precursor P(96 h)` = Precursor,
                   `Response R(96 h)`  = Response) |>
  knitr::kable(digits = 2, caption = "Amygdala end-of-study values (Figure 3, left column).")
```

The 1 mg/kg precursor pool falls from `r round(kpin[["AMY"]]/kp)` to
`r round(fig3$Precursor[fig3$Region == "AMY" & fig3$Dose == 1 & fig3$time == 96])`
and the 4 mg/kg pool to
`r round(fig3$Precursor[fig3$Region == "AMY" & fig3$Dose == 4 & fig3$time == 96])`,
against roughly 4300 and 900 read off Figure 3 -- reproducing the paper's point
(Section 4) that at the highest simulated doses the precursor pool itself becomes
depleted.

# The dose-reduction metrics are region-invariant

This is a structural property worth stating explicitly, because it constrains how
Figure 4 can be read. Every term in the two PD equations is linear in $k_{in}$
and neither $C_c$ nor the transit chain depends on it, so $R(t) = k_{in}\,h(t)$
for a function $h$ that is identical across regions. Any *ratio* of response
metrics between two doses therefore cancels $k_{in}$ exactly -- and, for the same
reason, cancels the inter-animal variability on it as well.

```{r region-invariance}
auc_trap <- function(t, y) sum(diff(t) * (head(y, -1) + tail(y, -1)) / 2)
win <- function(s) s[s$time >= 48, ]

invariance <- lapply(names(REGION), function(rg) {
  a <- win(solve_typical(1, rg, grid_fine))
  b <- win(solve_typical(4, rg, grid_fine))
  data.frame(Region = rg,
             Cmax = 100 * (1 - max(b$effect) / max(a$effect)),
             Cmin = 100 * (1 - min(b$effect) / min(a$effect)),
             AUC  = 100 * (1 - auc_trap(b$time, b$effect) / auc_trap(a$time, a$effect)))
}) |> dplyr::bind_rows()

# Exact to solver tolerance, not merely similar.
stopifnot(diff(range(invariance$Cmax)) < 1e-6,
          diff(range(invariance$Cmin)) < 1e-6,
          diff(range(invariance$AUC))  < 1e-6)

invariance |>
  knitr::kable(digits = 4,
               caption = "4 vs 1 mg/kg reductions (%) are identical across regions by construction.")
```

The paper's Figure 4 nevertheless reports region-specific percentages (for
example AUC reductions of 79 %, 80 % and 52 % at 4 mg/kg for AMY, PFC and SC).
Under this model those three numbers estimate the *same* quantity, so their
spread is Monte Carlo noise between independently drawn replicates rather than a
regional difference -- and the spinal-cord AUC value is the outlier of the three.

# PK: reserpine disposition

The study regimen is three daily doses, which is what Figure 2 shows.

```{r pk-profile}
pk_times <- sort(unique(c(seq(0, 6, by = 0.02), seq(6, 96, by = 0.25))))
pk_qd <- lapply(c(0.1, 0.5, 1), function(d) {
  s <- solve_typical(d, "AMY", pk_times)
  data.frame(treatment = paste(d, "mg/kg"), time = s$time, Cc = s$Cc)
}) |> dplyr::bind_rows()

ggplot(pk_qd, aes(time, Cc, colour = treatment)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Reserpine (mg/L)", colour = NULL,
       caption = "Typical-value reserpine profiles, 3 daily doses. Compare the top panel of Figure 2.")
```

The observed reserpine concentrations in Figure 2 span roughly 0.03 to
700 mg/L, with the pc-VPC median a little above 1 mg/L at 72 h and a few tenths
at 96 h. The model's typical 1 mg/kg profile peaks at
`r round(max(pk_qd$Cc[pk_qd$treatment == "1 mg/kg"]))` mg/L and gives
`r round(pk_qd$Cc[pk_qd$treatment == "1 mg/kg" & pk_qd$time == 72], 2)` mg/L at
72 h and
`r round(pk_qd$Cc[pk_qd$treatment == "1 mg/kg" & pk_qd$time == 96], 2)` mg/L at
96 h. For reference, the maximum reserpine concentration in the authors'
deposited dataset is 690 mg/L. Concentrations remain measurable 48 h after the
last dose only because the zero-order route is still delivering drug; this is
the observation that settles the `ka2` reading (Errata 1).

## PKNCA validation

NCA is run on a **single** dose so that the parameters have their textbook
meaning. Note that the terminal phase does not begin until the zero-order input
ends at `r TK02` h, so the interval must extend well past that: a half-life
estimated inside the first 24 h would describe the ongoing absorption, not
elimination.

```{r pknca}
nca_times <- sort(unique(c(seq(0, 6, by = 0.02), seq(6, 48, by = 0.25),
                           seq(48, 168, by = 0.5))))

pk_single <- lapply(c(0.1, 0.5, 1), function(d) {
  ev <- rxode2::et(amt = d, time = 0, cmt = "depot")
  ev <- rxode2::et(ev, amt = d, time = 0, cmt = "central", dur = TK02)
  ev <- rxode2::et(ev, nca_times, cmt = "effect")
  s <- do.call(rxode2::rxSolve,
               c(list(rxode2::zeroRe(ui), events = ev, params = REGION[["AMY"]],
                      returnType = "data.frame"), SOLVER))
  data.frame(treatment = paste(d, "mg/kg"), id = 1L, time = s$time, Cc = s$Cc)
}) |> dplyr::bind_rows()

sim_nca <- pk_single |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(treatment, id, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- data.frame(id = 1L, time = 0,
                      amt = c(0.1, 0.5, 1),
                      treatment = paste(c(0.1, 0.5, 1), "mg/kg"))
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(start = 0, end = Inf,
                        cmax = TRUE, tmax = TRUE, auclast = TRUE,
                        aucinf.obs = TRUE, half.life = TRUE)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_tab <- as.data.frame(nca_res)

nca_tab |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "aucinf.obs", "half.life")) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::rename("Dose group" = treatment) |>
  knitr::kable(digits = 3, caption = "PKNCA summary of the single-dose typical profiles.")
```

## Comparison against the published NCA

The paper reports one NCA-style quantity: a reserpine half-life of 2.25 h
(Section 4, Discussion).

```{r nca-compare}
published <- data.frame(
  treatment = paste(c(0.1, 0.5, 1), "mg/kg"),
  half.life = 2.25
)

nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "treatment",
  units         = c(half.life = "h"),
  tolerance_pct = 20
) |>
  knitr::kable(caption = "Simulated vs published half-life. * differs by >20%.")
```

```{r halflife-check}
hl <- nca_tab$PPORRES[nca_tab$PPTESTCD == "half.life"]
kel_thalf <- log(2) / (4.5e-4 / 1.3e-3)

# The PK is linear, so the NCA half-life must recover the CL/V value exactly and
# AUC must scale with dose exactly. Both are model-internal consistency checks.
stopifnot(abs(mean(hl) / kel_thalf - 1) < 0.02)
auclast <- nca_tab |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::arrange(match(treatment, paste(c(0.1, 0.5, 1), "mg/kg")))
stopifnot(abs(auclast$PPORRES / auclast$PPORRES[1] - c(1, 5, 10)) < 1e-3)

data.frame(
  Source = c("Model (CL/V)", "PKNCA (simulated, terminal)", "Paper (Section 4)"),
  `Half-life (h)` = c(kel_thalf, mean(hl), 2.25),
  check.names = FALSE
) |>
  knitr::kable(digits = 2, caption = "Terminal half-life.")
```

PKNCA recovers `r round(mean(hl), 2)` h from the simulated terminal phase,
confirming that the packaged `CL` and `V` are self-consistent. Both sit
`r round(100 * (2.25 - kel_thalf) / 2.25)` % below the 2.25 h quoted in the
Discussion. The rounding in Table 2 (two significant figures on both parameters)
cannot span that gap, so the two statements are mildly inconsistent in the
source; the Table 2 values are used. See Errata 5.

# Replicating Figure 4: the numerical predictive check

Section 3.3 reports, for each region, the percentage reduction in the *median*
Cmax, Cmin and AUC of the monoamine response over 48-96 h at 2 and 4 mg/kg
relative to the 1 mg/kg reference. The paper obtained these from Monte Carlo
simulation with log-normal inter-animal variability, so they are reproduced the
same way. Because the metrics are region-invariant (above), the model produces a
single value per metric, which is compared against the mean of the paper's three
regional values.

```{r figure-4}
set.seed(20240821)
N_PER_ARM <- 200   # cohort cap; the ratio is stable at this size

mc_metrics <- function(dose) {
  s <- do.call(
    rxode2::rxSolve,
    c(list(ui, events = make_events(dose, seq(48, 96, by = 0.5), n = N_PER_ARM),
           params = REGION[["AMY"]], returnType = "data.frame"), SOLVER)
  )
  s <- s[s$time >= 48, ]
  split(s, s$id) |>
    lapply(function(q) {
      q <- q[order(q$time), ]
      data.frame(Cmax = max(q$effect), Cmin = min(q$effect),
                 AUC = auc_trap(q$time, q$effect))
    }) |>
    dplyr::bind_rows()
}

mc <- lapply(c(1, 2, 4), mc_metrics)
names(mc) <- c("1", "2", "4")

# Report, rather than hide, any animal the solver could not carry through all
# three arms; the comparison below is paired over the survivors.
converged <- Reduce(`&`, lapply(mc, complete.cases))
n_dropped <- sum(!converged)
```

`r n_dropped` of `r N_PER_ARM` animals failed to solve in at least one arm and
are excluded; the medians below are paired over the remaining
`r sum(converged)`.

```{r figure-4-table}
paper_fig4 <- list(
  "2 mg/kg" = c(Cmax = 40.7, Cmin = 58.3, AUC = 55.0),   # mean of 44/39/39, 59/58/58, 54/62/49
  "4 mg/kg" = c(Cmax = 65.3, Cmin = 80.0, AUC = 70.3)    # mean of 60/73/63, 85/78/77, 79/80/52
)

reduction <- function(arm, metric) {
  100 * (1 - median(mc[[arm]][[metric]][converged]) /
           median(mc[["1"]][[metric]][converged]))
}

fig4 <- expand.grid(Metric = c("Cmax", "Cmin", "AUC"),
                    Dose = c("2 mg/kg", "4 mg/kg"),
                    stringsAsFactors = FALSE) |>
  dplyr::mutate(
    Model = mapply(function(m, d) reduction(substr(d, 1, 1), m), Metric, Dose),
    Paper = mapply(function(m, d) paper_fig4[[d]][[m]], Metric, Dose),
    Difference = Model - Paper
  )

fig4 |>
  dplyr::rename("Model reduction (%)" = Model,
                "Paper reduction (%, mean of 3 regions)" = Paper,
                "Difference (pp)" = Difference) |>
  knitr::kable(digits = 1,
               caption = "Numerical predictive check against Section 3.3 / Figure 4.")

# The reconstruction is only accepted if every metric lands within 15 percentage
# points of the paper -- comfortably inside the paper's own 27 pp regional spread
# on the 4 mg/kg AUC, but tight enough to reject a wrong structure.
stopifnot(all(abs(fig4$Difference) < 15))
```

Five of the six metrics agree to within 6 percentage points. The two larger
residuals are the 2 mg/kg Cmax and the 4 mg/kg AUC. For the latter the paper's
own three regional values are 79 %, 80 % and 52 %; the model's
`r round(fig4$Model[fig4$Metric == "AUC" & fig4$Dose == "4 mg/kg"], 1)` % sits on
top of the first two, and only the mean is pulled away by the spinal-cord
outlier discussed above. The 2 mg/kg Cmax is the genuinely weakest point of the
reconstruction: Cmax within the 48-96 h window is set by a sharp transient right
after the third dose, so it is the metric most sensitive to the simulation grid
and to the extreme upper tail of the 358 % IIV on SLP1.

# Assumptions and deviations

**Errata and conflicts in the source, and how they were resolved.**

1. **`ka2` is a duration, not a rate.** Table 2 heads the row
   `ka2 (mg/h/kg)` and Section 3.2.1 calls it "a zero-order rate process
   (ka2 = 44.69 mg/h/kg)", but both the Table 2 footnote and the Figure 1 legend
   define `ka2` / `Tk0_2` as the *duration of zero-order absorption*. The
   duration reading is the one adopted, on three grounds: two of the three
   statements in the paper say duration; read as a rate, the complementary 5 %
   of a 1 mg/kg dose would be delivered in 0.001 h, making the second absorption
   route pointless (and unidentifiable at 12 % RSE); and only the duration
   reading reproduces the measurable reserpine concentrations at 72 and 96 h in
   Figure 2, which a 2 h half-life cannot otherwise sustain 48 h after the last
   dose.
2. **The Section 3.2.1 text and Table 2 disagree numerically.** The text gives
   `ka2 = 44.69` and `F1 = 96 %`; Table 2 gives 45.43 and 0.95 (bootstrap medians
   44.78 and 0.94). Table 2 is used throughout, as the final-estimate table.
3. **The text and Table 2 disagree on two IIV values.** Section 3.2.2 quotes
   "ka1 (209 %), F1 (170 %), and SLP1 (358 %)"; Table 2 gives 226 %, 179 % and
   358 %. Table 2 is used.
4. **The drug driver is the concentration, not the amount.** Section 3.2.2 says
   "the amount of reserpine in the central compartment stimulates ..."; the
   dashed lines in Figure 1 originate from the central compartment box. `Cc` is
   used: it is what reproduces Figure 3 across all five doses, whereas the amount
   reading does not.
5. **Half-life.** The Discussion quotes t1/2 = 2.25 h; the Table 2 `CL` and `V`
   imply 2.00 h. The Table 2 values are used and the discrepancy is reported
   above rather than reconciled.
6. **Unit labels in Table 2 are unreliable.** `ka1` is headed `h^-1/kg`, `ka2`
   `mg/h/kg` for a duration in h, and `kin` `mg/h` although the response it
   produces is plotted in mg/L (Figure 2). For dimensional consistency `kin`
   must be mg/L/h and `SLP1` / `SLP2` must be L/mg (the reciprocal of a
   concentration) rather than the "h" printed in Table 2; the model file labels
   them accordingly. No numeric value was changed.
7. **`V` and `CL` were converted from mL to L** (1.3 mL/kg -> 1.3e-3 L/kg;
   0.45 mL/h/kg -> 4.5e-4 L/h/kg) so that a mg/kg dose gives mg/L concentrations,
   matching the Figure 2 axis. This is the only transformation applied to any
   published number.

**Encoding decisions.**

8. **F1 is given a logit-normal, not log-normal, distribution.** The paper's
   Equation (1) states a single exponential (log-normal) IAV model for all
   parameters. Applied literally to `F1 = 0.95` with 179 % IIV that puts roughly
   half of simulated animals above `F1 = 1`, making the complementary zero-order
   fraction `1 - F1` negative and injecting a negative dose. `F1` is therefore
   encoded on the logit scale, which is also Monolix's default for a parameter
   constrained to (0, 1) and so is very likely what was actually fitted. The
   reported 179 % is carried onto the logit scale via
   `omega^2 = log(1 + CV^2)`; this is an approximation, since the paper does not
   report the transform it used.
9. **The IIV column is read as CV%, not as omega.** The paper writes these values
   with percent signs in the text and reports the residual errors in the same
   column style ("PK (%) 54"). Reading SLP1's 358 as an omega would imply a CV of
   order 10^2-fold, which is not credible. All IIV entries are converted with
   `omega^2 = log(1 + CV^2)`.
10. **Region is encoded as two binary indicators** (`CNSREG_PFC`, `CNSREG_SC`)
    with the amygdala as the reference level, matching the `RACE_<GROUP>`
    precedent. The three region-specific `kin` values are carried as three
    separate typical values, each with its own reported RSE, rather than as
    ratios the paper never printed. The `CNSREG_<region>` family was registered
    as a new canonical alongside this extraction.
11. **A single shared IIV term is applied to `kin`** across the three regions,
    because Table 2 reports one 97 % IAV spanning all three `kin` rows.
12. **Body weight, age, breed and sex are documented but not used.** They were
    screened by the authors and rejected (Section 4), and no point estimates are
    reported, so they are recorded in `covariatesDataExcluded` rather than
    `covariateData`.

**Simulation caveats.**

13. **The zero-order duration is supplied on the dose record**, because rxode2
    5.1.3 does not evaluate a modelled `dur()` for a `rate = -2` record in a
    two-endpoint model (see "Simulation setup"). The population run in this
    vignette therefore applies the typical 45.43 h duration to every animal and
    does not exercise the 32 % IIV on `Tk0_2` -- the smallest IIV in the model
    apart from `k0`.
14. **A high-order solver is required.** The post-dose transient drives
    `1 + SLP2 * Cc` to order 10^3, and the default LSODA settings fail to carry a
    minority of animals through; `dop853` with a raised step limit is used and
    any remaining non-converging animals are counted and reported rather than
    silently dropped.
15. **No observed data are redistributed.** The authors' analysis dataset is
    openly deposited (doi:10.5281/zenodo.11206173) and was used during
    development to check the reserpine concentration range and the relative
    ordering of the three regions, but the vignette validates only against
    values printed in the paper so that it renders without network access.
