---
title: "Factor VIII (Allard 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Factor VIII (Allard 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Allard_2020_factorviii"))$reference`
* Description: `r rxode2::rxode(readModelDb("Allard_2020_factorviii"))$description`
* Article (open access): <https://doi.org/10.3390/pharmaceutics12040380>

Allard and colleagues pooled routine-care FVIII:C (one-stage clotting assay)
measurements after eight FVIII concentrates -- seven standard half-life (SHL)
products and one extended half-life (EHL) product, Elocta (efmoroctocog alfa,
rFVIIIFc) -- into a single Monolix population PK model. The final model is a
linear two-compartment model with power effects of weight and age on clearance,
weight on the central volume, and an Elocta indicator on clearance.

# Population

The analysis included 258 patients from 13 French haemophilia treatment centres
(87 children and adolescents, 171 adults; Allard 2020 Table 1): median age 30
years (3-77), median weight 64 kg (15.1-130). Haemophilia was severe in 244
patients, moderate in 11 and minor in 3; none had inhibitors. PK profiles were
recorded after Elocta (n = 136), Advate (44), Kogenate (34), Refacto (18),
Factane (8), Kovaltry (7), Novoeight (6) and Afstyla (5); 44 patients switched
from an SHL product to Elocta and contributed a profile on each. The 935 FVIII:C
samples (1-11 per patient, median 4) were taken before and 15 min to 96 h after
an intravenous dose of median 2750 IU (500-5000 IU); 10.6% were below the LLOQ.
Sex is not reported.

The same information is available programmatically via
`readModelDb("Allard_2020_factorviii")()$population`.

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` | log(204 mL/h) | Table 2, Cl |
| `lvc` | log(2640 mL) | Table 2, V1 |
| `lq` | log(135 mL/h) | Table 2, Q |
| `lvp` | log(339 mL) | Table 2, V2 |
| `e_wt_cl` | 0.75 (fixed) | Table 2, beta Weight on Cl '0.75 FIX'; Section 3.3 |
| `e_age_cl` | -0.214 | Table 2, beta Age; sign from the Table 2 footnote equation |
| `e_form_fviii_fc_cl` | -0.394 | Table 2, beta EHL; sign from the Table 2 footnote equation |
| `e_wt_vc` | 0.827 | Table 2, beta Weight on V1 |
| `etalcl` variance | 0.349^2 = 0.121801 | Table 2, omega Cl 34.9% |
| `etalvc` variance | 0.232^2 = 0.053824 | Table 2, omega V1 23.2% |
| cov(`etalcl`, `etalvc`) | 0.599 x 0.349 x 0.232 = 0.048500 | Table 2, Corr(V1, Cl) |
| `etalvp` variance | 0.468^2 = 0.219024 | Table 2, omega V2 46.8% |
| `addSd` | 0.00868 IU/mL | Table 2, a (constant) |
| `propSd` | 0.108 | Table 2, b (proportional) 10.8% |
| Continuous covariate model | `(COV / median)^beta` | Eq. 1; medians 30 years and 64 kg from the Table 2 footnote |
| Categorical covariate model | `exp(beta * EHL)` | Eq. 2 |
| Residual error | `y = Cc + (a + b * Cc) * e` | Eq. 3 (Monolix combined1) |
| Two-compartment linear ODEs | n/a | Section 3.2 |

# Closed-form checks of the typical patient

The terminal half-life of a two-compartment model follows from its
micro-constants. For the typical patient (30 years, 64 kg) the model gives:

```{r closed-form}
mod <- readModelDb("Allard_2020_factorviii")
ini_est <- rxode2::rxode2(mod)$theta

half_life_2cmt <- function(cl, vc, q, vp) {
  k10 <- cl / vc
  k12 <- q / vc
  k21 <- q / vp
  s <- k10 + k12 + k21
  beta <- (s - sqrt(s^2 - 4 * k10 * k21)) / 2
  log(2) / beta
}

cl_shl <- exp(ini_est[["lcl"]])
cl_ehl <- exp(ini_est[["lcl"]] + ini_est[["e_form_fviii_fc_cl"]])
vc_typ <- exp(ini_est[["lvc"]])
q_typ <- exp(ini_est[["lq"]])
vp_typ <- exp(ini_est[["lvp"]])

typical <- tibble(
  product = c("Standard half-life", "Elocta"),
  `CL (mL/h)` = c(cl_shl, cl_ehl),
  `Terminal t1/2 (h)` = c(
    half_life_2cmt(cl_shl, vc_typ, q_typ, vp_typ),
    half_life_2cmt(cl_ehl, vc_typ, q_typ, vp_typ)
  )
)
knitr::kable(typical, digits = 2)
hl_ratio_typ <- typical$`Terminal t1/2 (h)`[2] / typical$`Terminal t1/2 (h)`[1]
hl_ratio_typ
```

The paper reports that half-life increased on average by a factor of 1.48 when
patients switched from an SHL product to Elocta (Section 3.6). The model's
typical-patient ratio, `r round(hl_ratio_typ, 3)`, reproduces it. Because the
peripheral exchange is slow relative to elimination, the terminal half-life
scales almost exactly with 1/CL, so the ratio is close to exp(0.394) = 1.483.
A reversed sign on `e_form_fviii_fc_cl` would give a ratio of about 0.68.

```{r closed-form-check}
stopifnot(abs(hl_ratio_typ - 1.48) < 0.03)
```

The closed form must also agree with the simulated typical profile, whose late
log-linear slope is the terminal rate constant. Both sides use the same
parameters, so the tolerance is tight.

```{r closed-form-sim}
ev_typ <- bind_rows(
  tibble(id = 1L, time = 0, evid = 1L, amt = 2750, cmt = "central"),
  tibble(id = 1L, time = seq(0, 96, by = 0.5), evid = 0L, amt = 0, cmt = "central")
) |>
  mutate(WT = 64, AGE = 30, FORM_FVIII_FC = 0L)
ev_typ <- bind_rows(ev_typ, mutate(ev_typ, id = 2L, FORM_FVIII_FC = 1L))

sim_typ <- rxode2::rxSolve(rxode2::zeroRe(mod), ev_typ, returnType = "data.frame")
late <- sim_typ |> filter(time >= 48)
hl_sim <- late |>
  group_by(id) |>
  summarise(hl = -log(2) / coef(lm(log(Cc) ~ time))[["time"]], .groups = "drop")
hl_sim
stopifnot(all(abs(hl_sim$hl / typical$`Terminal t1/2 (h)` - 1) < 0.005))
```

# Virtual cohort

Original observed data are not public. The virtual cohort approximates Table 1:
one third children and adolescents (3-17 years) and two thirds adults (18-77
years), with weights tied to age so that the overall medians sit near 30 years
and 64 kg. As in the switch analysis of the paper, every virtual patient
receives both an SHL product and Elocta. To keep the pairing, the random effects
are drawn once per patient in R from the model's OMEGA and supplied as data
columns to the typical-value (`zeroRe()`) model, so each patient carries the
same random effects on both products.

```{r cohort}
set.seed(20200421)
n_sub <- 200L
n_child <- round(n_sub * 87 / 258)

child <- tibble(AGE = runif(n_child, 3, 17)) |>
  mutate(WT = exp(log(16) + 0.095 * (AGE - 3) + rnorm(n(), 0, 0.15)))
adult <- tibble(AGE = pmin(pmax(exp(rnorm(n_sub - n_child, log(37), 0.35)), 18), 77)) |>
  mutate(WT = exp(rnorm(n(), log(72), 0.18)))
cohort <- bind_rows(child, adult) |>
  mutate(WT = pmin(pmax(WT, 15.1), 130), subject = seq_len(n()))

om <- rxode2::rxode2(mod)$omega
eta_names <- c("etalcl", "etalvc", "etalvp")
om <- om[eta_names, eta_names]
etas <- matrix(rnorm(n_sub * 3), n_sub, 3) %*% chol(om)
colnames(etas) <- eta_names
cohort <- bind_cols(cohort, as_tibble(etas))

cohort |>
  summarise(
    `Median age (y)` = median(AGE), `Age range` = paste(round(range(AGE)), collapse = "-"),
    `Median weight (kg)` = median(WT), `Weight range` = paste(round(range(WT), 1), collapse = "-")
  ) |>
  knitr::kable(digits = 1)
```

Each patient receives a single 40 IU/kg intravenous dose (close to the median
2750 IU for a 64 kg patient) of each product, sampled to 96 h.

```{r events}
obs_times <- c(0, 0.25, 0.5, 1, 2, 4, 6, 8, 12, 24, 36, 48, 72, 96)
make_events <- function(cohort, form, id_offset) {
  base <- cohort |>
    mutate(id = subject + id_offset, FORM_FVIII_FC = form, dose = 40 * WT)
  doses <- base |> mutate(time = 0, evid = 1L, amt = dose, cmt = "central")
  obs <- base |>
    select(-dose) |>
    tidyr::crossing(time = obs_times) |>
    mutate(evid = 0L, amt = 0, cmt = "central")
  bind_rows(select(doses, -dose), obs) |> arrange(id, time, desc(evid))
}
events <- bind_rows(
  make_events(cohort, 0L, 0L),
  make_events(cohort, 1L, 1000L)
) |>
  mutate(product = ifelse(FORM_FVIII_FC == 1L, "Elocta", "Standard half-life"))

sim <- rxode2::rxSolve(
  rxode2::zeroRe(mod), events,
  keep = c("product", "subject", "AGE", "WT"),
  returnType = "data.frame"
)
```

# Concentration-time profiles

```{r profiles, fig.cap = "Simulated individual FVIII:C after 40 IU/kg on each product (median and 10th-90th percentile band); dashed lines mark 0.01 and 0.02 IU/mL troughs used in the paper's clinical example."}
sim |>
  group_by(product, time) |>
  summarise(
    med = median(Cc), lo = quantile(Cc, 0.1), hi = quantile(Cc, 0.9),
    .groups = "drop"
  ) |>
  filter(time > 0) |>
  ggplot(aes(time, med, colour = product, fill = product)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, colour = NA) +
  geom_line() +
  geom_hline(yintercept = c(0.01, 0.02), linetype = "dashed") +
  scale_y_log10() +
  labs(x = "Time after dose (h)", y = "FVIII:C (IU/mL)", colour = NULL, fill = NULL) +
  theme_bw()
```

# Half-life after switching (Figure 6)

Figure 6 of the paper compares, for the 44 switchers, the half-life on the SHL
product with the half-life on Elocta; the mean ratio was 1.48 and the multiple
linear regression in Section 3.6 attributed +4.8 h to the switch, +0.13 h per
year of age and -0.055 h per kg of weight.

```{r switch-hl, fig.cap = "Replicates Figure 6 of Allard 2020: paired terminal half-lives of each virtual patient on a standard half-life product and on Elocta."}
hl_ind <- cohort |>
  mutate(
    cl = exp(ini_est[["lcl"]] + etalcl) * (WT / 64)^0.75 * (AGE / 30)^ini_est[["e_age_cl"]],
    vc = exp(ini_est[["lvc"]] + etalvc) * (WT / 64)^ini_est[["e_wt_vc"]],
    vp = exp(ini_est[["lvp"]] + etalvp),
    hl_shl = half_life_2cmt(cl, vc, q_typ, vp),
    hl_ehl = half_life_2cmt(cl * exp(ini_est[["e_form_fviii_fc_cl"]]), vc, q_typ, vp),
    ratio = hl_ehl / hl_shl
  )

hl_ind |>
  select(subject, `Standard half-life` = hl_shl, Elocta = hl_ehl) |>
  pivot_longer(-subject, names_to = "product", values_to = "hl") |>
  mutate(product = factor(product, c("Standard half-life", "Elocta"))) |>
  ggplot(aes(product, hl, group = subject)) +
  geom_line(alpha = 0.2) +
  geom_point(alpha = 0.4) +
  labs(x = NULL, y = "Terminal half-life (h)") +
  theme_bw()

hl_ind |>
  summarise(
    `Median SHL t1/2 (h)` = median(hl_shl), `Median Elocta t1/2 (h)` = median(hl_ehl),
    `Mean ratio` = mean(ratio), `Median ratio` = median(ratio)
  ) |>
  knitr::kable(digits = 2)
```

```{r switch-regression}
reg_data <- bind_rows(
  hl_ind |> transmute(hl = hl_shl, switch = 0, AGE, WT),
  hl_ind |> transmute(hl = hl_ehl, switch = 1, AGE, WT)
)
fit <- lm(hl ~ switch + AGE + WT, data = reg_data)
knitr::kable(
  tibble(
    Term = c("Switch to Elocta (h)", "Age (h/year)", "Weight (h/kg)"),
    `Allard 2020 Section 3.6` = c(4.8, 0.13, -0.055),
    Simulated = unname(coef(fit)[c("switch", "AGE", "WT")])
  ),
  digits = 3
)
stopifnot(
  abs(median(hl_ind$ratio) - 1.48) < 0.05,
  coef(fit)[["switch"]] > 3, coef(fit)[["switch"]] < 7,
  coef(fit)[["AGE"]] > 0
)
```

With no within-patient (between-occasion) variability in the model, every
virtual patient's half-life is longer on Elocta; the four switchers (9%) whose
observed half-life fell after switching reflect occasion-to-occasion variation
and estimation noise in sparse routine data that the model does not describe.
The regression on the simulated patients reproduces the switch effect and the
direction of the age effect. The weight coefficient is not asserted: in the
paper it came from the 44 switchers only, and in the model the weight exponents
on CL (0.75) and V1 (0.827) nearly cancel in the half-life.

# PKNCA validation

NCA on the noise-free individual profiles, grouped by product.

```{r pknca}
conc <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, product)
dose_df <- events |>
  filter(evid == 1L) |>
  select(id, time, amt, product)

conc_obj <- PKNCAconc(conc, Cc ~ time | product + id)
dose_obj <- PKNCAdose(dose_df, amt ~ time | product + id)
intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE, cl.obs = TRUE
)
nca <- pk.nca(PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_summary <- as.data.frame(nca$result) |>
  filter(PPTESTCD %in% c("cmax", "aucinf.obs", "half.life", "cl.obs")) |>
  group_by(product, PPTESTCD) |>
  summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = PPTESTCD, values_from = median)
nca_summary |>
  dplyr::rename(
    Product = product,
    "Cmax (IU/mL)" = cmax,
    "AUC0-inf (IU*h/mL)" = aucinf.obs,
    "t1/2 (h)" = half.life,
    "CL (mL/h)" = cl.obs
  ) |>
  knitr::kable(digits = 3)
```

The paper does not tabulate NCA results. The one comparable figure is the
Discussion's median Elocta half-life of 14.8 h among the study patients (the
same model's individual estimates), which the simulated cohort median is
compared against below. The cohort is only an approximation of Table 1, so a
difference of up to 20% is expected.

```{r nca-comparison}
cmp <- ncaComparisonTable(
  simulated = as.data.frame(nca$result) |> filter(product == "Elocta"),
  reference = data.frame(half.life = 14.8),
  params = "half.life",
  units = c(half.life = "h")
)
knitr::kable(cmp)
hl_elocta <- nca_summary$half.life[nca_summary$product == "Elocta"]
stopifnot(abs(hl_elocta / 14.8 - 1) < 0.2)
```

# Clinical example (Figure 7)

Section 5 illustrates dose individualisation for a 21-year-old, 105 kg man on
Afstyla (a standard half-life product) dosed every 48 h. His individual
parameters (Figure 7A) are empirical Bayes estimates that are not printed, so
the typical-value profile for his covariates is shown instead of his own.

```{r clinical-example, fig.cap = "Typical-value FVIII:C for a 21-year-old, 105 kg patient on a standard half-life product at 1000 or 2000 IU every 48 h (compare Figure 7B-C of Allard 2020, which uses the patient's individual estimates)."}
ev_ex <- lapply(c(1000, 2000), function(d) {
  rxode2::et(amt = d, ii = 48, addl = 9, cmt = "central") |>
    rxode2::et(seq(0, 480, by = 1), cmt = "central") |>
    as.data.frame() |>
    mutate(id = ifelse(d == 1000, 1L, 2L), regimen = paste(d, "IU q48h"))
}) |>
  bind_rows() |>
  mutate(WT = 105, AGE = 21, FORM_FVIII_FC = 0L)
sim_ex <- rxode2::rxSolve(rxode2::zeroRe(mod), ev_ex, keep = "regimen", returnType = "data.frame")
sim_ex |>
  filter(time >= 384) |>
  ggplot(aes(time - 384, Cc, colour = regimen)) +
  geom_line() +
  geom_hline(yintercept = c(0.01, 0.02), linetype = "dashed") +
  scale_y_log10() +
  labs(x = "Time after dose at steady state (h)", y = "FVIII:C (IU/mL)", colour = NULL) +
  theme_bw()
sim_ex |>
  filter(time == 480) |>
  select(regimen, `Trough FVIII:C (IU/mL)` = Cc) |>
  knitr::kable(digits = 4)
```

The paper reports that this patient, with his individual estimates, stays
above 0.01 IU/mL at 48 h on 1000 IU (1% risk of falling below) and above
0.02 IU/mL for 48 h on 2000 IU. The typical patient with his covariates has
lower troughs than that, so his own clearance must be below the typical
value. That is the paper's argument for individual PK assessment, and the
typical-value profile is not expected to reproduce it.

# Assumptions and deviations

* **Signs of the age and Elocta coefficients.** Table 2 prints both as negative
  (-0.214 and -0.394), with negative bootstrap intervals, and the footnote
  equation `CLi = CLpop x (Age/30)^-0.214 x (weight/64)^0.75 x e^-0.394`
  confirms it. Both directions agree with the reported longer half-life on
  Elocta and in older patients.
* **Random-effect scale.** Monolix reports omega as the standard deviation of
  the random effect, so the Table 2 percentages are read as omega = 0.349,
  0.232 and 0.468 and squared to give the variances in `ini()`. The Corr(V1, Cl)
  row is a correlation coefficient, converted to a covariance. Q has no random
  effect ("fixed random effect" in the abstract).
* **Residual error.** Eq. 3 is the Monolix combined1 form, in which the
  additive and proportional standard deviations add
  (`y = Cc + (a + b * Cc) * e`), encoded with `combined1()`.
* **Endogenous and residual FVIII.** The authors fitted the data at steady
  state, adding four virtual prior doses at the known dosing interval to
  account for pre-dose (including endogenous) FVIII:C, and handled BLQ data as
  right-truncated. These are data-handling steps; the model has no endogenous
  production term (the authors tested one and did not retain it), so
  simulations start from zero FVIII:C.
* **Product indicator.** The paper's `EHL` covariate is encoded as
  `FORM_FVIII_FC`. Elocta was the only extended half-life product in the data,
  so the effect applies to the Fc-fusion product, not to other extended
  half-life FVIII concentrates.
* **Screened covariates.** von Willebrand factor, ABO blood group and fat-free
  mass were tested but not retained (missing for many patients); they are
  listed in `covariatesDataExcluded`. Their estimates are in Supplementary
  Table S2, which is not used by the final model.
* **Virtual cohort.** Age and weight distributions are approximations of
  Table 1; sex is not reported. No erratum was found in a Europe PMC search on
  2026-09-26.
