---
title: "CYP3A4 induction by rifampicin and other inducers (Almond 2016)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CYP3A4 induction by rifampicin and other inducers (Almond 2016)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Article: <https://doi.org/10.1124/dmd.115.066845>
* Erratum: <https://doi.org/10.1124/dmd.115.066845err>
* Supplement: open-access supplementary material deposited with PMC4885489

Almond and colleagues asked whether physiologically based dynamic modelling can
predict the magnitude of CYP3A drug-drug interactions caused by enzyme
*induction*. The study has two halves. The first is bench work: CYP3A4 induction
was characterised in cryopreserved human hepatocytes from four donors for six
inducers, measuring mRNA and catalytic activity in parallel. The second is a
Simcyp simulation exercise that used those parameters, plus in vivo reference
values for rifampicin, to predict 29 clinical rifampicin DDI studies and six
studies with other inducers.

**This package carries the first half.** Seven model files are shipped: six
in-vitro concentration-response models, one per inducer, and one model carrying
the refined *in vivo* rifampicin reference parameters that are the paper's
headline result. The section
["What is not reproducible"](#what-is-not-reproducible) below accounts in detail
for the Simcyp layer that is deliberately not carried.

```{r models}
invitro_models <- c(
  rifampicin    = "Almond_2016_rifampicin_invitro",
  carbamazepine = "Almond_2016_carbamazepine_invitro",
  phenobarbital = "Almond_2016_phenobarbital_invitro",
  phenytoin     = "Almond_2016_phenytoin_invitro",
  efavirenz     = "Almond_2016_efavirenz_invitro",
  nifedipine    = "Almond_2016_nifedipine_invitro"
)
invivo_model <- "Almond_2016_rifampicin_invivo"

# Resolve each model to an rxUi exactly once. readModelDb() returns the model
# FUNCTION, so it must be passed through rxode2::rxode() before any `$` access
# (pattern 7 of known-vignette-failure-patterns.md).
uis <- lapply(c(invitro_models, invivo = invivo_model),
              function(nm) rxode2::rxode(readModelDb(nm)))

cat(uis[["rifampicin"]]$reference)
```

# Population

The in-vitro models describe cryopreserved human hepatocytes from **four donors**
(Hu1206, Hu1191, Hu1198, Hu4193). Cells were incubated with serial dilutions of
each inducer, prepared daily in dimethyl sulfoxide, in culture medium containing
0.1% dimethyl sulfoxide (v/v). Two endpoints were measured **in parallel in the
same incubations**: CYP3A4 catalytic activity, as 6-beta-hydroxytestosterone
formation by LC-MS/MS, and CYP3A4 mRNA, by the QuantiGene Plex 2.0 assay. Cell
toxicity and viability were monitored by lactate dehydrogenase leakage and
AlamarBlue. Because the two readouts come from the same cells, each inducer is
one model with two outputs rather than two model files.

The in-vivo rifampicin model is anchored on 29 clinical DDI studies in white
subjects (Table 3 of the source), 10 with the victim drug given intravenously
(116 subjects) and 19 orally (241 subjects), covering midazolam, alfentanil,
alprazolam, nifedipine, simvastatin, zolpidem and triazolam. Its reference
induction parameters themselves were derived from the change in the
6-beta-hydroxycortisol / cortisol metabolic ratio during rifampicin 600 mg daily
for 14 days, combined with separately published rifampicin concentration-time
data.

The same information is available programmatically, e.g.
`readModelDb("Almond_2016_rifampicin_invitro")()$population`.

# The model

All seven files share one structural form, the three-parameter Emax
concentration-response that the source fitted in GraphPad Prism (Hill exponent
constrained to 1; the authors report that a four-parameter fit was "not
significantly different" and used the simpler model):

$$\text{fold}(C) = 1 + \frac{E_{max}\,C}{EC_{50} + C}$$

This is exactly the induction term of the paper's enzyme-turnover equations
(Eqs. 3 and 4), which read $1 + (Ind_{max}-1)\,I/(IndC_{50}+I)$.

**The single most important reading trap in this source** is that `Indmax` is
the maximum *fold* induction and is **not** baseline-corrected: the paper states
that it "is equal to Emax + 1". The shipped files therefore carry
`emax = Indmax - 1` and `ec50 = IndC50`, following the translation already
established in `Willemin_2024_interleukin6_cyp_pbpk.R` and
`Chen_2024_interleukin6_cyp3a_pbpk.R`. Section
["Calibration"](#calibration-eqs-5-and-6) shows that the paper's own calibration
equation is a pure ratio on `Indmax - 1`, which is independent confirmation that
`emax` is the natural parameterisation.

# Source trace

Every value below is also recorded as an in-file comment beside its `ini()`
entry. This table collects them for review.

```{r source-trace, echo = FALSE}
tribble(
  ~Quantity, ~Value, ~Source,
  "fold(C) = 1 + emax * C / (ec50 + C)", "n/a", "Eqs. 3 and 4 (induction term); Methods, 'In Vitro Data Analysis' (3-parameter fit, Hill = 1)",
  "Indmax = Emax + 1", "n/a", "Methods, 'In Vitro Data Analysis'",
  "Indmax / IndC50, six inducers, both endpoints", "24 values", "Table 2 (mean and S.D. over four donors)",
  "Incubation concentration ranges", "7 per inducer", "Table 1",
  "Hepatocyte donors", "Hu1206, Hu1191, Hu1198, Hu4193", "Methods, 'Materials'",
  "In vivo rifampicin IndC50", "0.32 uM", "Methods, 'Design of Virtual Studies'; Supplemental Table 2",
  "In vivo rifampicin Indmax, base model A", "8", "Methods, 'Design of Virtual Studies'; Supplemental Table 2",
  "In vivo rifampicin Indmax, refined model C", "16", "Methods, 'Design of Virtual Studies'; Table 5",
  "Calibration of Indmax", "Eq. 5", "Methods, 'Derivation of Reference In Vivo Induction Parameters'",
  "Calibration of IndC50", "Eq. 6", "Methods, 'Derivation of Reference In Vivo Induction Parameters'",
  "Model C prediction accuracy (GMFE 1.48, 79.3% within limits)", "n/a", "Table 5",
  "Molecular weights used for unit notes", "n/a", "Supplemental Tables 1 and 2"
) |>
  knitr::kable(caption = "Source trace for every shipped value and equation.")
```

# Shipped values against Table 2

The first validation is a transcription audit: pull the parameters back out of
the packaged models and confirm they reproduce Table 2 exactly.

```{r table2-audit}
# Published Table 2, transcribed independently of the model files.
published <- tribble(
  ~inducer,        ~Indmax_activity, ~SD_Indmax_activity, ~IndC50_activity, ~SD_IndC50_activity,
                   ~Indmax_mrna,     ~SD_Indmax_mrna,     ~IndC50_mrna,     ~SD_IndC50_mrna,
  "rifampicin",    22.7,  7.8,  0.30, 0.10,  29.9,  7.0,  0.71, 0.35,
  "carbamazepine", 16.6,  6.1, 59.1,  37.3,  21.9, 12.4, 58.7,  18.0,
  "phenobarbital", 21.1, 11.5, 473,  245,    44.2, 25.9, 743,   334,
  "phenytoin",     13.6,  3.7, 51.3,  29.4,  24.5,  7.6, 123,   120,
  "efavirenz",     13.5,  4.2,  4.9,   1.7,  18.1,  5.4,   8.4,   5.1,
  "nifedipine",    15.6, 11.3,  4.0,   1.9,  30.0, 22.0,  13.0,   9.5
)

# Typical values as shipped, recovered from ini().
theta_of <- function(ui) {
  d <- ui$iniDf
  d <- d[is.na(d$neta1), ]
  stats::setNames(d$est, d$name)
}

shipped <- lapply(names(invitro_models), function(nm) {
  th <- theta_of(uis[[nm]])
  tibble(
    inducer          = nm,
    Indmax_activity  = exp(th[["lemax_activity"]]) + 1,
    IndC50_activity  = exp(th[["lec50_activity"]]),
    Indmax_mrna      = exp(th[["lemax_mrna"]]) + 1,
    IndC50_mrna      = exp(th[["lec50_mrna"]])
  )
}) |> bind_rows()

audit <- published |>
  select(inducer, starts_with("Ind")) |>
  left_join(shipped, by = "inducer", suffix = c("_pub", "_ship"))

# Deterministic identity: these are the same constants read two ways, so an
# exact-to-floating-point comparison is the correct assertion here (no
# simulation, no random draw). Any transcription slip fails it immediately.
stopifnot(
  all(abs(audit$Indmax_activity_pub - audit$Indmax_activity_ship) < 1e-9),
  all(abs(audit$IndC50_activity_pub - audit$IndC50_activity_ship) < 1e-9),
  all(abs(audit$Indmax_mrna_pub     - audit$Indmax_mrna_ship)     < 1e-9),
  all(abs(audit$IndC50_mrna_pub     - audit$IndC50_mrna_ship)     < 1e-9)
)

audit |>
  transmute(
    Inducer                = inducer,
    `Indmax activity`      = Indmax_activity_ship,
    `IndC50 activity (uM)` = IndC50_activity_ship,
    `Indmax mRNA`          = Indmax_mrna_ship,
    `IndC50 mRNA (uM)`     = IndC50_mrna_ship
  ) |>
  knitr::kable(digits = 2,
               caption = "Shipped typical values; all 24 reproduce Table 2 exactly.")
```

The rifampicin row gets a second, independent confirmation from elsewhere in the
paper: Table 5 labels its uncalibrated models D and E with the very parameters
they were run at -- "Indmax 22.7, IndC50 0.30" for the activity model and
"Indmax 29.9, IndC50 0.71" for the mRNA model. Those are Table 2's rifampicin
values reprinted in a different table for a different purpose, so agreement
between the two is not circular.

```{r table5-crosscheck}
stopifnot(
  abs(shipped$Indmax_activity[shipped$inducer == "rifampicin"] - 22.7) < 1e-9,
  abs(shipped$IndC50_activity[shipped$inducer == "rifampicin"] - 0.30) < 1e-9,
  abs(shipped$Indmax_mrna[shipped$inducer == "rifampicin"]     - 29.9) < 1e-9,
  abs(shipped$IndC50_mrna[shipped$inducer == "rifampicin"]     - 0.71) < 1e-9
)
```

# The paper's own narrative as an answer key

The Results section states three summary ranges over the six inducers. Because
each is a ratio computed from Table 2, together they over-determine all 24
numbers: a single mis-transcribed value moves at least one ratio outside its
stated bound. This is the strongest free check available on this transcription,
and it needs no simulation.

> "efficacy was higher (1.3- to 2.0-fold higher Indmax values), but potency
> (IndC50) was generally lower (1.0- to 3.3-fold) when measured by changes in
> mRNA levels compared with changes in activity. When the ratio of Indmax to
> IndC50 was compared, no systematic trend was seen [...] with fold difference
> between the two ranging from 0.6- to 1.3-fold."

```{r narrative-answer-key}
ratios <- shipped |>
  transmute(
    Inducer                 = inducer,
    `Indmax mRNA/activity`  = Indmax_mrna / Indmax_activity,
    `IndC50 mRNA/activity`  = IndC50_mrna / IndC50_activity,
    `Indmax:IndC50 mRNA/activity` =
      (Indmax_mrna / IndC50_mrna) / (Indmax_activity / IndC50_activity)
  )

# Deterministic (no simulation). Bounds are the paper's own quoted ranges,
# widened only by the rounding implied by its one-decimal reporting: the paper
# says "1.3- to 2.0-fold" for values that compute to 1.32-2.09, "1.0- to
# 3.3-fold" for 0.99-3.25, and "0.6- to 1.3-fold" for 0.56-1.34.
stopifnot(
  all(ratios$`Indmax mRNA/activity`  >= 1.25),
  all(ratios$`Indmax mRNA/activity`  <= 2.15),
  all(ratios$`IndC50 mRNA/activity`  >= 0.95),
  all(ratios$`IndC50 mRNA/activity`  <= 3.35),
  all(ratios$`Indmax:IndC50 mRNA/activity` >= 0.55),
  all(ratios$`Indmax:IndC50 mRNA/activity` <= 1.35)
)

ratios |>
  knitr::kable(digits = 2,
               caption = paste("mRNA-versus-activity ratios. Observed spans:",
                               "Indmax 1.32-2.09 (paper: 1.3-2.0);",
                               "IndC50 0.99-3.25 (paper: 1.0-3.3);",
                               "ratio-of-ratios 0.56-1.34 (paper: 0.6-1.3)."))
```

All three reproduce. Carbamazepine defines the lower bound of the IndC50 range
(0.99), nifedipine the upper (3.25), and phenobarbital the upper Indmax bound
(2.09) -- so the check is genuinely exercised at both ends by three different
inducers, not satisfied trivially.

# Replicating Figure 2

Figure 2 of the source compares Indmax (panel A), IndC50 (panel B) and their
ratio (panel C) between the mRNA and activity readouts across the six inducers,
with lines of unity, 0.8-1.25-fold and 0.5-2-fold. Reproduced here from the
packaged parameters.

```{r figure-2, fig.height = 3.8}
# Panels A and B carry the published mean +/- S.D. error bars. Panel C cannot:
# the S.D. of the Indmax:IndC50 ratio would need the per-donor ratios, which the
# source does not publish (only per-parameter means and S.D.s).
fig2 <- bind_rows(
  shipped |>
    left_join(published, by = "inducer", suffix = c("", "_p")) |>
    transmute(inducer, panel = "A: Indmax (fold)",
              activity = Indmax_activity, mrna = Indmax_mrna,
              sd_activity = SD_Indmax_activity, sd_mrna = SD_Indmax_mrna),
  shipped |>
    left_join(published, by = "inducer", suffix = c("", "_p")) |>
    transmute(inducer, panel = "B: IndC50 (uM)",
              activity = IndC50_activity, mrna = IndC50_mrna,
              sd_activity = SD_IndC50_activity, sd_mrna = SD_IndC50_mrna),
  shipped |>
    transmute(inducer, panel = "C: Indmax:IndC50",
              activity = Indmax_activity / IndC50_activity,
              mrna     = Indmax_mrna / IndC50_mrna,
              sd_activity = NA_real_, sd_mrna = NA_real_)
)

ggplot(fig2, aes(activity, mrna)) +
  geom_abline(slope = 1, intercept = 0) +
  geom_abline(slope = c(0.8, 1.25), intercept = 0, linetype = "dotted") +
  geom_abline(slope = c(0.5, 2), intercept = 0, linetype = "dashed") +
  geom_errorbar(aes(ymin = pmax(mrna - sd_mrna, 1e-3), ymax = mrna + sd_mrna),
                width = 0, alpha = 0.45, na.rm = TRUE) +
  geom_errorbarh(aes(xmin = pmax(activity - sd_activity, 1e-3),
                     xmax = activity + sd_activity),
                 height = 0, alpha = 0.45, na.rm = TRUE) +
  geom_point(aes(shape = inducer), size = 2.4) +
  facet_wrap(~panel, scales = "free") +
  scale_x_log10() + scale_y_log10() +
  scale_shape_manual(values = c(16, 15, 17, 3, 4, 8)) +
  labs(x = "Activity-derived value", y = "mRNA-derived value", shape = NULL,
       caption = "Replicates Figure 2 of Almond 2016 (mean +/- S.D. over four donors).") +
  theme(legend.position = "bottom")
```

Panel C shows the paper's point directly: the Indmax:IndC50 ratios straddle the
line of unity and all sit inside the 0.5-2-fold band, i.e. no systematic trend
between the two readouts once efficacy and potency are combined.

# Structural checks on the concentration-response curve

Three identities follow from the functional form and must hold exactly for every
model. They are asserted against the *same drawn parameters* used in the solve,
so the difference is pure floating-point error and a tight bound is the correct
assertion (a loose one here would hide a real error).

```{r structural-checks}
# Solve each in-vitro model at C = 0, C = ec50, and a saturating C, using the
# typical values (zeroRe()). No ODE states in these models, so observation rows
# need no `cmt`.
check_one <- function(nm) {
  ui  <- uis[[nm]]
  th  <- theta_of(ui)
  cov <- paste0("CP_", switch(nm,
                              rifampicin    = "RIF",
                              carbamazepine = "CARBAMAZEPINE",
                              phenobarbital = "PHENOBARBITAL",
                              phenytoin     = "PHENYTOIN",
                              efavirenz     = "EFAVIRENZ",
                              nifedipine    = "NIFEDIPINE"), "_UM")
  e50a <- exp(th[["lec50_activity"]])
  e50m <- exp(th[["lec50_mrna"]])
  emaxa <- exp(th[["lemax_activity"]])
  emaxm <- exp(th[["lemax_mrna"]])

  ev <- data.frame(id = 1L, time = seq_len(4) - 1, evid = 0)
  ev[[cov]] <- c(0, e50a, e50m, 1e7 * max(e50a, e50m))

  # `uis[[nm]]` is the already-resolved rxUi: zeroRe() and rxSolve() both
  # require an rxUi, and readModelDb() returns the model FUNCTION. `omega = NA`
  # is the explicit no-random-effects sentinel -- zeroRe() alone is NOT enough,
  # because rxode2 keeps the previous solve's omega in the compiled model's
  # solve options and would otherwise re-sample etas here.
  out <- rxode2::rxSolve(rxode2::zeroRe(ui), ev, omega = NA,
                         returnType = "data.frame")
  tibble(
    inducer         = nm,
    fold0_activity  = out$foldActivity[1],
    fold0_mrna      = out$foldMrna[1],
    half_activity   = out$foldActivity[2] - (1 + emaxa / 2),
    half_mrna       = out$foldMrna[3]     - (1 + emaxm / 2),
    sat_activity    = out$foldActivity[4] - (1 + emaxa),
    sat_mrna        = out$foldMrna[4]     - (1 + emaxm)
  )
}

struct <- bind_rows(lapply(names(invitro_models), check_one))

stopifnot(
  # fold(0) = 1 exactly: no inducer, no induction.
  all(abs(struct$fold0_activity - 1) < 1e-12),
  all(abs(struct$fold0_mrna     - 1) < 1e-12),
  # fold(ec50) = 1 + emax/2: this IS the definition of the reported IndC50,
  # "the concentration that yields half of the Emax".
  all(abs(struct$half_activity) < 1e-8),
  all(abs(struct$half_mrna)     < 1e-8),
  # fold(C -> Inf) = 1 + emax = Indmax, approached to within 1e-5 fold at the
  # saturating concentration used here.
  all(abs(struct$sat_activity) < 1e-5),
  all(abs(struct$sat_mrna)     < 1e-5)
)

struct |>
  transmute(Inducer = inducer,
            `fold(0)` = fold0_activity,
            `fold(ec50) - (1+emax/2), activity` = half_activity,
            `fold(ec50) - (1+emax/2), mRNA`     = half_mrna,
            `fold(sat) - Indmax, activity`      = sat_activity) |>
  knitr::kable(digits = 12, caption = "Structural identities; all residuals are numerical noise.")
```

# Concentration-response curves over the assayed ranges

Typical-value curves over each inducer's own Table 1 concentration ladder. The
assayed points are marked, which makes visible how well each curve's plateau is
actually pinned by the experiment -- rifampicin's top concentration is 100-fold
above its IndC50, while phenobarbital's is only about 4-fold above its mRNA
IndC50.

```{r curves, fig.height = 5}
assayed <- list(
  rifampicin    = c(0.03, 0.1, 0.3, 1, 3, 10, 30),
  carbamazepine = c(1, 3, 10, 30, 100, 300, 1000),
  phenobarbital = c(10, 30, 100, 300, 1000, 2000, 3000),
  phenytoin     = c(1, 3, 10, 30, 100, 300, 1000),
  efavirenz     = c(0.1, 0.3, 1, 2, 3, 10, 30),
  nifedipine    = c(0.03, 1, 2, 3, 10, 30, 100)
)

covname <- c(rifampicin = "CP_RIF_UM", carbamazepine = "CP_CARBAMAZEPINE_UM",
             phenobarbital = "CP_PHENOBARBITAL_UM", phenytoin = "CP_PHENYTOIN_UM",
             efavirenz = "CP_EFAVIRENZ_UM", nifedipine = "CP_NIFEDIPINE_UM")

curve_one <- function(nm) {
  rng  <- range(assayed[[nm]])
  grid <- 10^seq(log10(rng[1]) - 0.5, log10(rng[2]) + 0.5, length.out = 120)
  ev <- data.frame(id = 1L, time = seq_along(grid) - 1, evid = 0)
  ev[[covname[[nm]]]] <- grid
  out <- rxode2::rxSolve(rxode2::zeroRe(uis[[nm]]), ev, omega = NA,
                         returnType = "data.frame")
  tibble(inducer = nm, conc = grid,
         Activity = out$foldActivity, mRNA = out$foldMrna) |>
    pivot_longer(c(Activity, mRNA), names_to = "endpoint", values_to = "fold")
}

curves <- bind_rows(lapply(names(invitro_models), curve_one))

points <- bind_rows(lapply(names(invitro_models), function(nm) {
  ev <- data.frame(id = 1L, time = seq_along(assayed[[nm]]) - 1, evid = 0)
  ev[[covname[[nm]]]] <- assayed[[nm]]
  out <- rxode2::rxSolve(rxode2::zeroRe(uis[[nm]]), ev, omega = NA,
                         returnType = "data.frame")
  tibble(inducer = nm, conc = assayed[[nm]],
         Activity = out$foldActivity, mRNA = out$foldMrna) |>
    pivot_longer(c(Activity, mRNA), names_to = "endpoint", values_to = "fold")
}))

ggplot(curves, aes(conc, fold, colour = endpoint)) +
  geom_line() +
  geom_point(data = points, size = 1.3) +
  facet_wrap(~inducer, scales = "free_x") +
  scale_x_log10() +
  labs(x = "Inducer concentration in culture medium (uM)",
       y = "CYP3A4 fold induction over vehicle control", colour = NULL,
       caption = "Typical-value curves; points mark the Table 1 assay concentrations.") +
  theme(legend.position = "bottom")
```

# Between-donor variability

Table 2 reports a standard deviation over the four donor-level fits. The shipped
models encode that as log-normal between-donor variability, so a simulated donor
cohort should recover the published central values and a spread of the right
order.

```{r between-donor}
n_donors <- 200L  # cohort cap is 200 per arm

donor_sim <- function(nm) {
  ev <- data.frame(id = seq_len(n_donors), time = 0, evid = 0)
  ev[[covname[[nm]]]] <- 0
  # Population run: pass `omega` EXPLICITLY. The typical-value solves above ran
  # zeroRe() on these same compiled models, and rxode2 retains that solve's
  # all-zero omega in the compiled model's solve options -- so without this the
  # 200 "donors" silently collapse onto one typical donor. The omega-recovery
  # assertion below is the mechanical guard on that failure.
  out <- rxode2::rxSolve(uis[[nm]], ev, omega = uis[[nm]]$omega,
                         returnType = "data.frame")
  tibble(inducer = nm,
         emax_activity = out$emax_activity, ec50_activity = out$ec50_activity,
         emax_mrna = out$emax_mrna,         ec50_mrna = out$ec50_mrna)
}

donors <- bind_rows(lapply(names(invitro_models), donor_sim))

# Realised central value and realised spread per inducer.
#
# The spread is asserted on the LOG-SCALE standard deviation, which estimates
# omega directly. That choice is deliberate and load-bearing. The arithmetic CV
# of a strongly right-skewed log-normal is dominated by the few largest draws,
# so its sampling distribution has a long upper tail that does NOT shrink
# usefully at n = 200: over 20,000 simulated cohorts drawn from this exact omega
# set, the realised-to-published arithmetic-CV ratio reached 3.6, and a band as
# wide as [0.6, 1.6] still failed about 1% of the time. Because rxode2 draws a
# different cohort per solver-thread count, an assertion that fails 1% of the
# time locally fails intermittently in CI (pattern 12 of
# known-vignette-failure-patterns.md). The log-scale SD has relative standard
# error 1/sqrt(2n) = 5% REGARDLESS of omega and is not tail-dominated, so it
# supports a TIGHTER band (25%, i.e. 5 standard errors) than the arithmetic CV
# supports at any width. The arithmetic CV is still reported below for
# comparison against the published S.D.; it is simply not asserted on.
realised <- donors |>
  group_by(inducer) |>
  summarise(across(c(emax_activity, ec50_activity, emax_mrna, ec50_mrna),
                   list(med = ~ stats::median(.x),
                        om  = ~ stats::sd(log(.x)),
                        cv  = ~ stats::sd(.x) / mean(.x))),
            .groups = "drop")

# Expected typical value and omega RE-DERIVED INDEPENDENTLY from published
# Table 2 -- not read back out of the model files. omega^2 = log(1 + CV^2), with
# the efficacy CV taken on the SHIFTED mean emax = Indmax - 1. This makes the
# check a test of the whole encoding chain rather than a tautology, and in
# particular it is the only automated check that exercises the shift-on-CV rule,
# which is the easiest thing to get wrong in this extraction.
pub <- published |>
  transmute(inducer,
            emax_activity_typ   = Indmax_activity - 1,
            ec50_activity_typ   = IndC50_activity,
            emax_mrna_typ       = Indmax_mrna - 1,
            ec50_mrna_typ       = IndC50_mrna,
            emax_activity_ompub = sqrt(log(1 + (SD_Indmax_activity / (Indmax_activity - 1))^2)),
            ec50_activity_ompub = sqrt(log(1 + (SD_IndC50_activity / IndC50_activity)^2)),
            emax_mrna_ompub     = sqrt(log(1 + (SD_Indmax_mrna / (Indmax_mrna - 1))^2)),
            ec50_mrna_ompub     = sqrt(log(1 + (SD_IndC50_mrna / IndC50_mrna)^2)))

cmp  <- left_join(realised, pub, by = "inducer")
pars <- c("emax_activity", "ec50_activity", "emax_mrna", "ec50_mrna")

med_err <- unlist(lapply(pars, function(p) abs(cmp[[paste0(p, "_med")]] / cmp[[paste0(p, "_typ")]]   - 1)))
om_err  <- unlist(lapply(pars, function(p) abs(cmp[[paste0(p, "_om")]]  / cmp[[paste0(p, "_ompub")]] - 1)))

stopifnot(
  # All six inducers x four parameters actually got compared -- guards against a
  # join or a name change silently reducing this to a check of nothing.
  length(med_err) == 24L, length(om_err) == 24L,
  !anyNA(med_err), !anyNA(om_err),
  # Median recovery. The median of a log-normal is its typical value; this is a
  # loose first layer (the exact transcription audit above is the strict one),
  # widened to sit clear of the sampling noise of the largest omega here (0.82).
  max(med_err) < 0.40,
  # Omega recovery: 5 standard errors, and the tight check of the two.
  max(om_err) < 0.25
)

cmp |>
  transmute(Inducer = inducer,
            `omega emax activity (sim)`     = emax_activity_om,
            `omega emax activity (Table 2)` = emax_activity_ompub,
            `omega ec50 mRNA (sim)`         = ec50_mrna_om,
            `omega ec50 mRNA (Table 2)`     = ec50_mrna_ompub,
            `arith. CV ec50 mRNA (sim)`     = ec50_mrna_cv) |>
  knitr::kable(digits = 2,
               caption = paste("Simulated between-donor log-scale SD against the omega implied by",
                               "the Table 2 S.D. (n = 200 simulated donors). The final column shows",
                               "the arithmetic CV, reported but not asserted on."))
```

Note that a log-normal CV of about 0.98 -- phenytoin's mRNA IndC50, where the
published S.D. of 120 uM nearly equals the mean of 123 uM -- produces a strongly
right-skewed donor distribution. With only four real donors behind it, that
parameter should be read as weakly determined rather than as a precise estimate
of population spread.

# The in vivo rifampicin model

The paper's headline result is that the Simcyp V12 base model, with rifampicin
`Indmax = 8`, systematically **underpredicted** the interaction when the victim
drug was given orally, and that raising `Indmax` to 16 in both liver and gut
(model C) fixed it.

```{r invivo-table, echo = FALSE}
tribble(
  ~Model, ~`Indmax (liver / gut)`, ~`IndC50 (uM)`, ~`GMFE, all victims`, ~`% within acceptance limits`,
  "A (base, Simcyp V12)", "8 / 8",   0.32, 2.12, 48.3,
  "B",                    "8 / 16",  0.32, 1.77, 51.7,
  "C (shipped)",          "16 / 16", 0.32, 1.48, 79.3,
  "F",                    "12 / 12", 0.32, 1.63, 79.3,
  "G",                    "20 / 20", 0.32, 1.51, 75.9
) |>
  knitr::kable(caption = "Table 5 of Almond 2016: prediction accuracy by rifampicin model. Lower GMFE is better.")
```

`Almond_2016_rifampicin_invivo` carries model C. Model B is the one alternative
the shipped single-parameter file cannot express, because it uses different
`Indmax` values in liver and gut; representing it needs two enzyme pools, which
in turn needs the tissue-specific `kdeg` values the paper never reports.

```{r invivo-checks}
# This model has NO etas, so it needs neither zeroRe() nor `omega = NA`
# (pattern 9 of known-vignette-failure-patterns.md). Passing `omega = NA` to an
# eta-free model is an error on the released rxode2 (5.1.6 -- "invalid 'times'
# argument"); it is tolerated only on the unreleased 5.1.7. The in vitro solves
# above DO keep `omega = NA` because those models carry etas.
ev_iv <- data.frame(id = 1L, time = 0:3, evid = 0,
                    CP_RIF_UM = c(0, 0.32, 5, 1e6))
iv <- rxode2::rxSolve(uis[["invivo"]], ev_iv,
                      returnType = "data.frame")

stopifnot(
  abs(iv$foldCyp3a4[1] - 1)    < 1e-12,   # no rifampicin -> uninduced
  abs(iv$foldCyp3a4[2] - 8.5)  < 1e-8,    # fold(ec50) = 1 + emax/2 = 8.5
  abs(iv$foldCyp3a4[4] - 16)   < 1e-4     # saturating -> Indmax = 16
)

# The Discussion states that at clinical dosing the free rifampicin
# concentration "exceeded the IndC50 for rifampicin (0.32 uM) across most of the
# dosing interval". At 5 uM -- well inside the range the paper describes -- the
# curve is already close to its plateau, which is why the paper found little
# sensitivity to fugut.
tibble(`CP_RIF_UM (uM)` = ev_iv$CP_RIF_UM,
       `fold CYP3A4 induction` = iv$foldCyp3a4,
       `% of Indmax` = 100 * (iv$foldCyp3a4 - 1) / 15) |>
  knitr::kable(digits = 3, caption = "In vivo rifampicin induction curve at key concentrations.")
```

The in-vitro and in-vivo rifampicin **potencies** agree closely -- IndC50 0.30 uM
in hepatocytes against 0.32 uM in vivo, a 7% difference. The **efficacies** do
not: 22.7-fold (activity) and 29.9-fold (mRNA) in vitro against 16-fold in vivo.
That mismatch is the entire motivation for the calibration approach below.

# Calibration (Eqs. 5 and 6) {#calibration-eqs-5-and-6}

The paper scales another inducer's in-vitro parameters by the ratio of
rifampicin's in-vivo to in-vitro values, measured in the *same* assay:

$$Ind_{max,cal} = \left[\frac{Ind_{max,test}-1}{Ind_{max,RIF}-1}\times\left(Ind_{max,RIF\,in\,vivo}-1\right)\right]+1
\qquad
IndC_{50,cal} = \frac{IndC_{50,test}}{IndC_{50,RIF}}\times IndC_{50,RIF\,in\,vivo}$$

Written in the shipped parameterisation, Eq. 5 collapses to a pure ratio,
`emax_cal = emax_test / emax_RIF * emax_RIF_invivo` -- which is why `emax`, not
`Indmax`, is the natural parameter to carry.

```{r calibration}
rif <- shipped |> filter(inducer == "rifampicin")
iv_th <- theta_of(uis[["invivo"]])
emax_rif_invivo <- exp(iv_th[["lemax"]])   # 15
ec50_rif_invivo <- exp(iv_th[["lec50"]])   # 0.32

calibrate <- function(tbl, endpoint) {
  emax_col  <- paste0("Indmax_", endpoint)
  ec50_col  <- paste0("IndC50_", endpoint)
  emax_rif  <- rif[[emax_col]] - 1
  ec50_rif  <- rif[[ec50_col]]
  tibble(
    inducer  = tbl$inducer,
    endpoint = endpoint,
    Indmax_cal = (tbl[[emax_col]] - 1) / emax_rif * emax_rif_invivo + 1,
    IndC50_cal = tbl[[ec50_col]] / ec50_rif * ec50_rif_invivo
  )
}

calibrated <- bind_rows(calibrate(shipped, "activity"), calibrate(shipped, "mrna"))

# SELF-CONSISTENCY IDENTITY. Calibrating rifampicin against itself must return
# the in vivo reference values exactly, on BOTH endpoints -- the in-vitro terms
# cancel. This is deterministic algebra, so an exact bound is correct, and it
# simultaneously checks the in-vitro rifampicin file, the in-vivo file and the
# calibration implementation against each other.
self <- calibrated |> filter(inducer == "rifampicin")
stopifnot(
  nrow(self) == 2L,
  all(abs(self$Indmax_cal - (emax_rif_invivo + 1)) < 1e-9),
  all(abs(self$IndC50_cal - ec50_rif_invivo) < 1e-9)
)

calibrated |>
  pivot_wider(names_from = endpoint, values_from = c(Indmax_cal, IndC50_cal)) |>
  transmute(Inducer = inducer,
            `Indmax cal (activity)` = Indmax_cal_activity,
            `IndC50 cal (activity, uM)` = IndC50_cal_activity,
            `Indmax cal (mRNA)` = Indmax_cal_mrna,
            `IndC50 cal (mRNA, uM)` = IndC50_cal_mrna) |>
  knitr::kable(digits = 2, caption = paste(
    "In vitro parameters calibrated to the refined rifampicin reference",
    "(Indmax 16, IndC50 0.32 uM) via Eqs. 5 and 6. DERIVED here, not",
    "transcribed: the source reports only the resulting prediction accuracy",
    "(Table 6), not the calibrated parameter values themselves."))
```

The paper's assessment of these calibrations is in its Table 6, over six DDI
studies with carbamazepine, phenytoin and phenobarbital:

```{r table6, echo = FALSE}
tribble(
  ~Approach, ~`GMFE (activity)`, ~`GMFE (mRNA)`, ~`% within limits (activity)`, ~`% within limits (mRNA)`,
  "Uncalibrated",          1.39, 1.44, 83.3, 83.3,
  "Calibrated to Indmax 8",  1.68, 1.46, 33.3, 83.3,
  "Calibrated to Indmax 16", 1.49, 1.35, 66.7, 83.3
) |>
  knitr::kable(caption = "Table 6 of Almond 2016: prediction accuracy for non-rifampicin inducers.")
```

The authors' conclusion is deliberately hedged, and worth repeating because it
cuts against a naive reading of the calibration idea: calibrating mRNA data to
the refined rifampicin reference gave the best mRNA result (GMFE 1.35), but for
activity data the *uncalibrated* values performed best (GMFE 1.39 versus 1.49).
Their recommendation is therefore that "it is prudent to fully characterize in
vitro induction with prototypical inducers" when using an uncalibrated approach,
rather than that calibration is always the right move.

# What is not reproducible {#what-is-not-reproducible}

The paper's simulation half is a Simcyp Population-Based Simulator exercise and
is **not** carried by any model file. The accounting:

* **The enzyme-turnover ODEs (Eqs. 3 and 4) themselves.** The functional form is
  published in full, but the rate constants are not. The paper reports only that
  a sensitivity analysis was run on "different values of first-order rate
  constants (kdegH and kdegG) that describe endogenous turnover of active enzyme
  in the liver and gut" -- it never prints them. Neither are the basal enzyme
  amounts `Enz0,H-3A4` / `Enz0,G-3A4`. Substituting a Simcyp library default
  would make the shipped model unsourceable, so the shipped files carry the
  induction term alone, which is the steady-state induction ratio the full
  equation relaxes toward. A user who supplies their own `kdeg` can wrap it,
  following `Chen_2024_interleukin6_cyp3a_pbpk.R`.
* **The whole-body disposition models.** Supplemental Tables 1 and 2 are
  unusually generous compound cards -- MW, fu, B:P, logP, pKa, fa, ka, Qgut,
  fugut, minimal-PBPK Vss, and enzyme kinetics -- but the physiology needed to
  turn a bottom-up `CLint` into an `L/h` is absent: no CYP3A4 abundance, no
  MPPGL, no liver weight, no organ blood flows, no tissue volumes.
* **The virtual populations.** Subjects were generated by a correlated Monte
  Carlo procedure inside the simulator; the demographic and physiological
  covariance structure is not published.
* **The victim-drug exposure predictions** of Figs. 3-5 and Tables 5-6 follow
  from the two items above and so are likewise not reproducible here. They are
  quoted in this vignette as published numbers, never recomputed.

# Assumptions and deviations

* **Log-normal between-donor variability is an extraction choice.** Table 2
  reports an arithmetic mean and standard deviation over four donors and states
  no distributional form. The shipped models convert each to a log-normal
  variance, `omega^2 = log(1 + CV^2)`. With `n = 4`, the standard deviations are
  themselves imprecise; treat the encoded variability as indicative.
* **The CV for the efficacy parameters is computed on `emax`, not `Indmax`.**
  Because `emax = Indmax - 1` is a shift, the published S.D. carries over
  unchanged while the mean drops by 1. Using the unshifted CV would understate
  every efficacy variance. Example: rifampicin activity, S.D. 7.8 on Indmax
  22.7, gives CV = 7.8 / 21.7 = 0.359, not 7.8 / 22.7 = 0.344.
* **No residual error is reported anywhere in the source.** Each donor's curve
  was fitted in GraphPad Prism and only point estimates and between-donor S.D.
  are given; there is no residual model, no replicate-level variance and no
  goodness-of-fit statistic. Every residual-error term is `fixed(0)`, so the
  models return deterministic curves per donor.
* **No Hill exponent is carried.** This is the authors' own model-selection
  decision, not an extraction assumption: they fitted both three- and
  four-parameter sigmoidal models, found the parameters "not significantly
  different", and used the three-parameter fit with the Hill exponent at 1.
* **The in-vivo model's parameters are `fixed()`.** `Indmax = 16` was selected
  from the discrete set {8, 12, 16, 20} by comparing DDI prediction accuracy,
  not estimated by regression, and `IndC50 = 0.32 uM` was inherited unchanged
  from the base model. Neither carries a standard error in the source.
* **Supplemental Table 2 label typo.** The rifampicin base values appear on two
  adjacent rows labelled "CYP3A4 Indmax (Fold, Emax +1) 8" and "CYP3A5 IndC50
  (uM) 0.32". The second label is a typographical error for CYP3A4: the main
  text gives the base reference values as "Indmax = 8; IndC50 = 0.32" and states
  that "only CYP3A4 was considered as CYP3A5 induction is less well
  characterized". The value 0.32 uM is used as the CYP3A4 IndC50 on the
  authority of the main text.
* **Erratum.** The published erratum (Drug Metab Dispos 2016;44(6):877) corrects
  the units at the top of the sixth and seventh columns of Table 3 from
  `ng/L.h` to `ng/mL.h`. Table 3 carries the clinical DDI exposures against
  which the rifampicin models were selected. No value in any shipped model file
  is taken from Table 3, so no shipped parameter changes; the correction matters
  for reading those exposures, and both citations are recorded in every model
  file's `reference` field.
* **Efavirenz and nifedipine have no in-vivo verification in this source.**
  Both were characterised in vitro (Tables 1 and 2) but appear in none of the
  clinical DDI simulations, so their shipped parameters were never confronted
  with in vivo data by the authors. Nifedipine additionally has the widest
  between-donor spread in the study.
* **Observation names are non-canonical, deliberately.**
  `checkModelConventions()` flags `foldCyp3a4` in the single-output in-vivo
  model as not canonical (`Cc` is expected for drug-concentration outputs). The
  endpoint here is a dimensionless fold-induction ratio, not a concentration;
  naming it `Cc` would misstate what the model observes. This follows the
  precedent set by `Mukker_2026_tuvusertib_hERG.R`, whose `hergInh` output
  carries the same accepted warning.
* **`CP_RIF_UM` is reused for three distinct senses of "rifampicin
  concentration".** The register entry was founded on an instantaneous plasma
  concentration; here it additionally carries an in-vitro culture-medium
  concentration and an operational *unbound* concentration at the site of
  enzyme interaction. The quantity and units are identical in all three, but the
  values are not interchangeable -- supplying a total plasma concentration to
  the in-vivo induction model would overstate the driver by roughly `1 / fu`,
  and rifampicin `fu = 0.15`. Each model's `covariateData` entry states which
  sense it requires.
