---
title: "Micafungin in patients with and without cancer (Alqahtani 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Micafungin in patients with and without cancer (Alqahtani 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Alqahtani 2021 fitted a two-compartment model to micafungin plasma
concentrations from hospitalised adults with and without cancer and reported a
complete parameter set for each group (Table 2). The library ships the two
columns of Table 2 as two model files:

* `Alqahtani_2021_micafungin_cancer`: patients with cancer (n = 10).
* `Alqahtani_2021_micafungin_noncancer`: patients without cancer (n = 9).

```{r load}
mods <- c(
  cancer = "Alqahtani_2021_micafungin_cancer",
  noncancer = "Alqahtani_2021_micafungin_noncancer"
)
uis <- lapply(mods, function(nm) rxode2::rxode(readModelDb(nm)))
```

* Citation: `r uis$cancer$reference`
* Description (cancer): `r uis$cancer$description`
* Description (non-cancer): `r uis$noncancer$description`
* Article: <https://doi.org/10.3390/antibiotics10111363> (open access; Supplementary
  File S1 describes the error and random-effect model forms)

# Population

The prospective single-centre study at King Saud University Medical City
(Riyadh, Saudi Arabia) enrolled 19 hospitalised adults who received at least two
doses of micafungin, empirically or for a confirmed fungal infection: 10 with
cancer and 9 without (Alqahtani 2021 Section 3.1, Table 1). Mean age was 47.3
and 51.1 years, mean body weight 63.4 and 69.8 kg, and 40% and 33% were female
in the cancer and non-cancer groups respectively. Mean SOFA scores were 7 and 8;
AST, ALT, albumin, bilirubin and creatinine clearance did not differ
significantly between the groups. All cancer patients received 100 mg/day;
seven non-cancer patients received 100 mg/day and two received 150 mg/day, as
60-min IV infusions. Seven samples were drawn per patient at 1, 2, 4, 6, 8, 12
and 24 h after the start of the infusion (133 samples).

```{r population}
str(uis$cancer$population[c("n_subjects", "age_range", "weight_range", "sex_female_pct", "dose_range")])
str(uis$noncancer$population[c("n_subjects", "age_range", "weight_range", "sex_female_pct", "dose_range")])
```

# Source trace

| Equation / parameter | Cancer | Non-cancer | Source location |
|---|---|---|---|
| Two-compartment model, linear elimination from central | -- | -- | Section 3.2 |
| `lcl` (CL, L/h) | log(1.2) | log(0.6) | Table 2 |
| `lvc` (V1, L) | log(10.7) | log(12) | Table 2 |
| `lq` (Q, L/h) | log(0.144) | log(0.188) | Table 2 |
| `lvp` (V2, L) | log(3.5) | log(2.77) | Table 2 |
| `etalcl` (IIV CL) | 34.1% CV | 11.8% CV | Table 2; log-normal per Supplementary File S1 |
| `etalvc` (IIV V1) | 7.6% CV | 7.6% CV | Table 2 |
| `etalq` (IIV Q) | 32.2% CV | 20.4% CV | Table 2 |
| `etalvp` (IIV V2) | 36.8% CV | 32.1% CV | Table 2 |
| `addSd` (residual `a`, mg/L) | 0.21 | 0.15 | Table 2 |
| `propSd` (residual `b`) | 0.22 | 0.18 | Table 2 |
| `Cobs = Cpred (1 + eps_prop) + eps_const` (combined2) | -- | -- | Supplementary File S1 |
| 60-min infusion of 100-150 mg/day | -- | -- | Section 2.2 |

The IIV rows are coefficients of variation (Table 2 footnote), converted to
log-normal variances as `omega^2 = log(1 + CV^2)`.

# Typical-value profiles

```{r typical}
ev_typ <- rxode2::et(amt = 100, dur = 1, ii = 24, addl = 13, cmt = "central") |>
  rxode2::et(seq(0, 336, by = 0.25), cmt = "central")

typ <- bind_rows(lapply(names(mods), function(g) {
  s <- rxode2::rxSolve(rxode2::zeroRe(uis[[g]]), ev_typ, returnType = "data.frame")
  s$group <- g
  s
}))

ggplot(typ, aes(time / 24, Cc, colour = group)) +
  geom_line() +
  labs(
    x = "Time (days)", y = "Micafungin concentration (mg/L)", colour = NULL,
    title = "Typical-value profiles, 100 mg once daily (60-min infusion)"
  ) +
  theme_bw()
```

With these parameters the terminal half-life is about 6 h in the cancer group
and about 14 h in the non-cancer group; nearly all of the difference comes from
CL (1.2 vs 0.6 L/h), which is the only parameter Table 2 reports as differing
significantly between the groups.

# Stochastic simulation and PKNCA

200 virtual patients per group receive 100 mg once daily for 14 days. The last
dosing interval (312-336 h) is sampled densely for NCA.

```{r sim}
n_per_arm <- 200
obs_t <- c(seq(0, 312, by = 24), 312 + c(0.25, 0.5, 0.75, 1, 1.25, 1.5, 2, 3, 4, 6, 8, 10, 12, 16, 20, 24))
ev_sim <- rxode2::et(amt = 100, dur = 1, ii = 24, addl = 13, cmt = "central") |>
  rxode2::et(obs_t, cmt = "central") |>
  rxode2::et(id = seq_len(n_per_arm))

rxode2::rxSetSeed(20211108)
sim <- bind_rows(lapply(names(mods), function(g) {
  s <- rxode2::rxSolve(uis[[g]], ev_sim,
    returnType = "data.frame",
    rtol = 1e-10, atol = 1e-12
  )
  s$group <- g
  s
})) |>
  mutate(id = paste(group, id, sep = "-"))
stopifnot(nrow(sim) > 0, !anyNA(sim$Cc))
```

```{r vpc, fig.cap = "Simulated steady-state dosing interval (day 14), median and 5th-95th percentiles of the typical-plus-IIV concentration. Compare with the pooled prediction-corrected VPC of Alqahtani 2021 Figure 2."}
last_int <- sim |>
  filter(time >= 312) |>
  mutate(tad = time - 312) |>
  group_by(group, tad) |>
  summarise(
    q05 = quantile(ipredSim, 0.05), q50 = median(ipredSim), q95 = quantile(ipredSim, 0.95),
    .groups = "drop"
  )
ggplot(last_int, aes(tad, q50, colour = group, fill = group)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.2, colour = NA) +
  geom_line() +
  labs(x = "Time after start of infusion (h)", y = "Micafungin concentration (mg/L)", colour = NULL, fill = NULL) +
  theme_bw()
```

PKNCA computes the steady-state AUC over the last interval, per subject and
grouped by population.

```{r pknca}
conc_nca <- sim |>
  filter(!is.na(Cc), time >= 312) |>
  mutate(tad = time - 312, Cc = pmax(ipredSim, 0)) |>
  select(id, group, tad, Cc)
dose_nca <- conc_nca |>
  distinct(id, group) |>
  mutate(tad = 0, amt = 100)

conc_obj <- PKNCA::PKNCAconc(conc_nca, Cc ~ tad | group + id)
dose_obj <- PKNCA::PKNCAdose(dose_nca, amt ~ tad | group + id)
intervals <- data.frame(start = 0, end = 24, auclast = TRUE, cmax = TRUE, tmax = TRUE, cmin = TRUE)
nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_wide <- as.data.frame(nca$result) |>
  select(group, id, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

nca_wide |>
  group_by(group) |>
  summarise(
    across(c(cmax, cmin, auclast), ~ signif(median(.x), 3)),
    tmax = median(tmax)
  ) |>
  rename(
    "Group" = group, "Cmax,ss (mg/L)" = cmax, "Cmin,ss (mg/L)" = cmin,
    "AUC0-24,ss (mg*h/L)" = auclast, "Tmax (h)" = tmax
  ) |>
  knitr::kable(caption = "Simulated median steady-state NCA, 100 mg once daily.")
```

The paper reports no NCA table, so there is no published row to place beside
these values. At steady state `AUC0-24 = Dose / CL` for each simulated subject;
checking it against each subject's own drawn `cl` tests dose, infusion, unit and
volume handling together.

```{r auc-identity}
cl_i <- sim |>
  group_by(id) |>
  summarise(cl = first(cl), .groups = "drop")
chk <- nca_wide |>
  left_join(cl_i, by = "id") |>
  mutate(pct_diff = 100 * (auclast * cl / 100 - 1))
stopifnot(nrow(chk) == 2 * n_per_arm, !anyNA(chk$pct_diff))
chk |>
  group_by(group) |>
  summarise(median_pct = median(pct_diff), p90_abs_pct = quantile(abs(pct_diff), 0.9)) |>
  knitr::kable(digits = 3, caption = "Per-subject AUC0-24,ss x CL / Dose - 1 (%).")
# The residual is trapezoid error across the infusion peak plus any subject not
# fully at steady state after 13 doses; a mis-scaled dose, volume or CL moves
# the whole distribution by tens of percent.
stopifnot(
  abs(median(chk$pct_diff)) < 1,
  quantile(abs(chk$pct_diff), 0.9) < 2
)
```

# Probability of target attainment

The paper's targets are total-plasma `AUC0-24/MIC` of 3000 for non-parapsilosis
*Candida* spp. and 285 for *C. parapsilosis* (Section 2.7). With linear PK the
steady-state AUC is `Dose / CL_i` and `CL_i` is log-normal, so the PTA is
deterministic:

`PTA = Phi( (log(Dose / (target * MIC)) - log(CL_pop)) / omega_CL )`.

```{r pta}
mics <- c(0.002, 0.004, 0.008, 0.016, 0.032, 0.064, 0.125, 0.25, 0.5, 1, 2, 4)
pta_analytic <- function(ui, dose, target, mic) {
  th <- ui$theta
  om <- sqrt(ui$omega["etalcl", "etalcl"])
  100 * stats::pnorm((log(dose / (target * mic)) - th[["lcl"]]) / om)
}
pta <- expand.grid(
  group = names(mods), dose = c(100, 150, 200), mic = mics,
  organism = c("Candida spp.", "C. parapsilosis"), stringsAsFactors = FALSE
) |>
  mutate(
    target = ifelse(organism == "Candida spp.", 3000, 285),
    pta = mapply(function(g, d, t, m) pta_analytic(uis[[g]], d, t, m), group, dose, target, mic)
  )

ggplot(pta, aes(factor(mic), pta,
  colour = group, linetype = factor(dose),
  group = interaction(group, dose)
)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 90, linetype = "dotted") +
  facet_wrap(~organism, ncol = 1) +
  labs(
    x = "MIC (mg/L)", y = "PTA (%)", colour = NULL, linetype = "Dose (mg/day)",
    title = "Analytic steady-state PTA (compare Alqahtani 2021 Figures 3 and 4)"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The simulated cohort gives the same PTA as the closed form, which checks the
formula against the model rather than against itself.

```{r pta-sim}
pta_cell <- function(g, d, org, m) {
  v <- pta$pta[pta$group == g & pta$dose == d & pta$organism == org & pta$mic == m]
  if (length(v) != 1L) stop("no unique PTA cell for ", g, " ", d, " ", org, " ", m)
  v
}
sim_pta <- 100 * mean(chk$auclast[chk$group == "cancer"] / 0.032 >= 3000)
ana_pta <- pta_cell("cancer", 100, "Candida spp.", 0.032)
c(simulated = sim_pta, analytic = ana_pta)
# Binomial SE at n = 200 and p = 0.33 is 3.3 points; 12 is > 3 SE.
stopifnot(abs(sim_pta - ana_pta) < 12)
```

## Comparison with the published anchors

The Discussion states that 100 mg/day in patients with cancer achieved an
approximate PTA of 30% at an MIC of 0.032 mg/L for non-parapsilosis *Candida*
spp.; Figure 4 shows 95% at 0.016 mg/L for the same arm.

```{r pta-anchor}
anchor <- data.frame(
  Claim = c("Cancer, 100 mg, Candida spp., MIC 0.032", "Cancer, 100 mg, Candida spp., MIC 0.016"),
  Published = c(30, 95),
  Model = c(pta_cell("cancer", 100, "Candida spp.", 0.032), pta_cell("cancer", 100, "Candida spp.", 0.016))
)
knitr::kable(anchor, digits = 1, caption = "PTA (%): Discussion / Figure 4 vs the packaged cancer model.")
stopifnot(abs(anchor$Model - anchor$Published) < 6)
```

Table 3 reports, for each dose and a 70 kg patient, the highest MIC with a PTA
of at least 90%. The packaged files carry no body-weight effect (see
Assumptions), so the model breakpoint is compared with the 70 kg rows.

```{r breakpoints}
bp <- pta |>
  group_by(group, dose, organism) |>
  summarise(model_bp = max(c(0, mic[pta >= 90])), .groups = "drop")
table3_70kg <- tribble(
  ~group, ~dose, ~organism, ~published_bp,
  "cancer", 100, "Candida spp.", 0.016,
  "cancer", 100, "C. parapsilosis", 0.125,
  "cancer", 150, "Candida spp.", 0.032,
  "cancer", 150, "C. parapsilosis", 0.25,
  "cancer", 200, "Candida spp.", 0.032,
  "cancer", 200, "C. parapsilosis", 0.25,
  "noncancer", 100, "Candida spp.", 0.032,
  "noncancer", 100, "C. parapsilosis", 0.25,
  "noncancer", 150, "Candida spp.", 0.032,
  "noncancer", 150, "C. parapsilosis", 0.25,
  "noncancer", 200, "Candida spp.", 0.032,
  "noncancer", 200, "C. parapsilosis", 0.5
)
bp_cmp <- table3_70kg |>
  left_join(bp, by = c("group", "dose", "organism")) |>
  mutate(match = model_bp == published_bp)
stopifnot(nrow(bp_cmp) == 12, !anyNA(bp_cmp$model_bp))
bp_cmp |>
  rename(
    "Group" = group, "Dose (mg/day)" = dose, "Organism" = organism,
    "Table 3, 70 kg (mg/L)" = published_bp, "Model (mg/L)" = model_bp, "Match" = match
  ) |>
  knitr::kable(caption = "90% PTA breakpoints: Alqahtani 2021 Table 3 (70 kg) vs the packaged models.")

# Deterministic: the cancer model reproduces 5 of 6 cells; the non-cancer
# model 1 of 6 (see the discussion below). A changed CL, IIV or target moves
# these counts.
stopifnot(
  sum(bp_cmp$match[bp_cmp$group == "cancer"]) == 5,
  sum(bp_cmp$match[bp_cmp$group == "noncancer"]) == 1
)
```

The cancer model reproduces the published breakpoints except at 150 mg for
*Candida* spp. (model 78.7% PTA at 0.032 mg/L; Figure 4 shows 92%). The
non-cancer model is consistently one doubling dilution more optimistic than
Table 3. Its Table 2 IIV on CL is only 11.8% CV, so the model's PTA curve drops
steeply around the MIC at which `Dose / (0.6 L/h)` crosses the target; the
published non-cancer curves in Figures 3 and 4 are much flatter (for example
60% and 65% at 0.064 mg/L for 100 and 150 mg), which implies considerably more
between-subject spread in AUC than the printed CL IIV alone. The likeliest
source is the retained covariate model (body weight, AST and ALT on CL), whose
coefficients and reference values are not printed and are therefore not in the
packaged files. The published figures are also not fully consistent with each
other: fitting a log-normal AUC distribution to Figure 4's cancer 100 mg points
gives a median CL near 1.2 L/h, while Figure 3's cancer 100 mg points give a
median CL near 1.0 L/h, and the cancer 150 and 200 mg curves in Figure 4 are
closer together than linear PK allows. No parameter was adjusted to close
these gaps.

# Assumptions and deviations

* **Covariate effects not encoded.** The paper states that AST, ALT and body
  weight significantly influenced CL, and that BMI, body weight, total bilirubin
  and albumin affected V1 (Section 3.2), and Table 3 varies body weight
  (50/70/100 kg). No coefficient, functional form, or reference value for any of
  these effects is printed in the article or in Supplementary File S1, which
  gives only the generic `CL_j = CL_pop x exp(eta_j)` form. The packaged files
  therefore carry the Table 2 typical values without covariate effects; the
  covariates are recorded under `covariatesDataExcluded` in each file. The
  covariate reference values to which the Table 2 typical values correspond are
  also not stated.
* **Two files for one analysis.** The Methods say the two groups were
  co-modelled, but Table 2 prints a separate estimate and RSE for every fixed
  effect, IIV and residual-error term in each group. Group-specific IIV and
  residual error cannot be expressed through a single covariate-switched
  model, so each column of Table 2 is a separate file.
* **IIV scale.** Table 2 gives IIV as a coefficient of variation; the variance is
  `log(1 + CV^2)`.
* **Residual error.** Supplementary File S1 writes
  `Cobs = Cpred (1 + eps_prop) + eps_const` with independent proportional and
  constant errors, i.e. the combined2 form. Table 2 `a` is taken as the
  additive SD in mg/L and `b` as the proportional SD.
* **PTA timing.** The paper does not say on which day `AUC0-24` was evaluated.
  The steady-state AUC is used here; it reproduces the Discussion's 30% anchor.
  Residual error is not included in the PTA.
* **Units in Table 1.** Table 1 prints serum creatinine in mmol/L (means of 74.7
  and 63.6), which is only plausible as umol/L, and gives no units for albumin,
  AST, ALT or bilirubin; the population metadata reproduces the values as
  printed.
