---
title: "Tramadol and O-desmethyltramadol in older patients (Al-Qurain 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tramadol and O-desmethyltramadol in older patients (Al-Qurain 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
rxode2::rxSetSeed(20220215)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("AlQurain_2022_tramadol"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
cat("* Article: <https://doi.org/10.1007/s13318-022-00756-x>\n", sep = "")
```

Al-Qurain et al. fitted tramadol and its active CYP2D6-formed metabolite
O-desmethyltramadol (ODT) jointly in 21 older hospital inpatients taking
extended-release oral tramadol, using Monolix 2020 (SAEM). The final estimates
are in Table 3 of the paper. The open-access supplement holds the
structural-model and covariate-screening tables (Tables S1-S7) but no model code
and no covariate equation, so several encoding choices below rest on the
reasoning laid out in [Assumptions and deviations](#assumptions-and-deviations).

A search on 2026-10-02 (Europe PMC, and the Crossref record for the DOI) found
no erratum or correction notice for this article.

# Population

`r ui$population$n_subjects` patients aged 69-93 years (mean 83.2, SD 7.5) were
recruited at the Royal Adelaide Hospital (Australia) between June and November
2019 (Table 1). 43% were male; body weight was 33-112 kg (mean 68.3). All had
normal liver function tests. Renal function was wide-ranging: Cockcroft-Gault
creatinine clearance (CrCL) 16-351 mL/min (mean 65.1, SD 69.1). The
Identification of Seniors At Risk (ISAR) frailty score was 1-5 (mean 3, SD 1.2),
and 38% took a CYP3A4 inhibitor and 38% a CYP2D6 inhibitor. Doses were 25 mg
(n = 3), 50 mg (n = 6), 75 mg (n = 1) or 100 mg (n = 11). Up to five
finger-prick volumetric absorptive microsamples were taken per patient within
12 h of a dose (99 samples in total, all above the 10 ng/mL limit of
quantitation).

The same information is available programmatically via
`readModelDb("AlQurain_2022_tramadol")()$population`.

# Model structure

Figure 1 of the paper: oral tramadol is absorbed first-order (`ka`) into a
central compartment (`V1/F`) that exchanges with a peripheral compartment
(`V2/F`, `Q`) and is cleared linearly by `CL/F`. ODT is formed at the
first-order rate `Kt` from the tramadol central amount into a one-compartment
pool that shares the tramadol central volume and is cleared by `CLm/F`.

`CL/F` is the **total** apparent tramadol clearance, and the formation term does
not deplete tramadol. Methods 2.4.2 states that the fraction of tramadol
metabolised to ODT (`Fm`) is not identifiable from parent-only oral dosing, and
that only `Vm/(F km)` and `CLm/(F km)` can be determined; the ODT pool is
therefore an apparent amount scaled by `1/Fm`. This reading reproduces the
paper's steady-state VPC (Figure 6) and the CrCL dependence of Figure 11, both
shown below; a mass-balance reading in which `Kt` also drains tramadol reaches
only about half of the Figure 6 medians and leaves CrCL almost without effect
(see [Assumptions and deviations](#assumptions-and-deviations)).

Covariates enter as uncentred exponentials:

* `Q = 42.6 * exp(0.255 * ISAR)` L/h
* `CL/F = 6.04 * exp(0.00498 * CrCL)` L/h
* `V2/F = 379 * exp(0.0119 * CrCL)` L

```{r model-states}
# The model defines a cl / vc pair, which rxode2 could replace with an analytic
# linCmt() solution, discarding the metabolite compartment. Confirm the four ODE
# states are present.
stopifnot(identical(
  rxode2::rxState(rxode2::rxModelVars(ui)),
  c("depot", "central", "peripheral1", "central_m1")
))
```

# Source trace

| Quantity | Value in model | Source |
|---|---|---|
| `ka` | 2.96 1/h | Table 3, final model, 'K a, /h' |
| `V1/F` (`vc`) | 373 L | Table 3, 'V 1 / F , l' = 0.373 (x 1000; see Assumptions) |
| `Q` | 42.6 L/h at ISAR 0 | Table 3, 'Q , l/h' = 0.0426 (x 1000) |
| ISAR effect on Q | 0.255 | Table 3, 'ISAR effect on Q' |
| `V2/F` (`vp`) | 379 L at CrCL 0 | Table 3, 'V 2 / F , l' = 0.379 (x 1000) |
| CrCL effect on V2/F | 0.0119 per mL/min | Table 3, 'CrCL effect on V 2 / F' |
| `CL/F` (`cl`) | 6.04 L/h at CrCL 0 | Table 3, 'CL/ F , l/h' = 0.00604 (x 1000) |
| CrCL effect on CL/F | 0.00498 per mL/min | Table 3, 'CrCL effect on CL/F' |
| `Kt` (`kmet`) | 0.0492 1/h | Table 3, 'K t , l/h' (rate constant per footnote) |
| `CLm/F` (`cl_m1`) | 143 L/h | Table 3, 'CL m / F , l/h' = 0.143 (x 1000) |
| IIV (SD) ka, V1, Q, V2, CL, Kt, CLm | 0.378, 0.738, 0.0219, 0.974, 0.0661, 0.785, 0.216 | Table 3 'Omega' rows; footnote: presented as SD |
| Tramadol proportional error | 0.157 | Table 3, 'Tramadol, b (%)' |
| ODT additive error | 5.19 ng/mL | Table 3, 'O -desmethyltramadol, a (%)' |
| ODT proportional error | 0.189 | Table 3, 'O -desmethyltramadol, b (%)' |
| Structure (2-cmt parent, 1-cmt metabolite, first-order absorption and formation) | `model()` ODEs | Figure 1; Methods 2.4.2; Results 3.3 |
| Formation does not deplete tramadol (Fm-scaled ODT pool) | `d/dt(central)` has no `kmet` term | Methods 2.4.2 identifiability statement; checked against Figures 6 and 11 |
| Log-normal IIV | `exp(theta + eta)` | Methods 2.4.1, Eq. 1 |
| Covariate form | uncentred `exp(beta * COV)` | Not printed; Monolix default, checked against Figure 10 |
| Error models | proportional (tramadol), combined (ODT) | Results 3.3 |

# Typical-value checks

The covariate model is algebraic, so the typical individual parameters at any
covariate value follow exactly from `ini()`:

```{r typical-params}
mod_typ <- ui |> rxode2::zeroRe()
ini_vals <- ui$theta
typ <- data.frame(
  SCORE_ISAR = c(0, 3, 6, 3),
  CRCL = c(0, 65.1, 65.1, 120)
) |>
  dplyr::mutate(
    Q = exp(ini_vals[["lq"]] + ini_vals[["e_score_isar_q"]] * SCORE_ISAR),
    CL = exp(ini_vals[["lcl"]] + ini_vals[["e_crcl_cl"]] * CRCL),
    V2 = exp(ini_vals[["lvp"]] + ini_vals[["e_crcl_vp"]] * CRCL),
    t_half_terminal = log(2) * (exp(ini_vals[["lvc"]]) + V2) / CL
  )
typ |>
  dplyr::rename(
    "ISAR" = SCORE_ISAR,
    "CrCL (mL/min)" = CRCL,
    "Q (L/h)" = Q,
    "CL/F (L/h)" = CL,
    "V2/F (L)" = V2,
    "Approximate terminal t1/2 (h)" = t_half_terminal
  ) |>
  knitr::kable(digits = 3)

# The covariate terms reproduce the cohort-mean values quoted in the
# covariateData notes (CL/F 8.3 L/h and V2/F 820 L at CrCL 65.1 mL/min).
stopifnot(
  abs(typ$CL[2] - 8.35) < 0.05,
  abs(typ$V2[2] / 822 - 1) < 0.01
)
```

The approximate terminal half-life, `ln 2 (V1 + V2) / CL`, is long (several
days) because `CL/F` is small relative to the combined volume. Steady state is
therefore reached only after weeks of dosing, which the simulations below allow
for.

At steady state ODT follows tramadol with its short own half-life, so the
typical ODT/tramadol concentration ratio is `Kt x V1 / CLm` (the two species
share the same volume):

```{r typical-ratio}
ratio_typ <- exp(ini_vals[["lkmet"]]) * exp(ini_vals[["lvc"]]) / exp(ini_vals[["lcl_m1"]])
ratio_typ
```

# Replicate Figure 10 (typical profiles by ISAR)

Figure 10 shows typical single-dose tramadol profiles at ISAR 0, 3 and 6 after
50 and 100 mg, with the dose given at 1 h on the figure axis. The CrCL used is
not stated; the cohort mean 65.1 mL/min is used here (the single-dose peak is
insensitive to CrCL, see Figure 11 below).

```{r fig10}
fig10_subj <- tidyr::crossing(
  SCORE_ISAR = c(0, 3, 6),
  dose_mg = c(50, 100)
) |>
  dplyr::mutate(id = dplyr::row_number(), CRCL = 65.1)

fig10_ev <- dplyr::bind_rows(
  fig10_subj |>
    dplyr::mutate(time = 1, amt = dose_mg, evid = 1L, cmt = "depot", dvid = NA_integer_),
  tidyr::crossing(fig10_subj, time = seq(0, 12, by = 0.05)) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

fig10_sim <- rxode2::rxSolve(
  mod_typ,
  events = fig10_ev,
  keep = c("SCORE_ISAR", "dose_mg"),
  useLinCmt = FALSE
) |>
  as.data.frame()

ggplot(fig10_sim, aes(time, Cc, colour = factor(dose_mg))) +
  geom_line() +
  facet_wrap(~ paste("ISAR =", SCORE_ISAR)) +
  scale_colour_manual(values = c("50" = "black", "100" = "red")) +
  labs(
    x = "Time (h)", y = "Tramadol concentration (ng/mL)", colour = "Dose (mg)",
    caption = "Replicates Figure 10 of Al-Qurain 2022."
  )
```

Peak concentrations after 100 mg, against values read off Figure 10 by the
maintainers:

```{r fig10-table}
fig10_cmax <- fig10_sim |>
  dplyr::filter(dose_mg == 100) |>
  dplyr::group_by(SCORE_ISAR) |>
  dplyr::summarise(
    Cmax = max(Cc),
    C12 = Cc[abs(time - 12) < 1e-6],
    .groups = "drop"
  ) |>
  dplyr::mutate(
    Cmax_fig10 = c(240, 220, 195),
    C12_fig10 = c(90, 112, 115),
    pct_diff = 100 * (Cmax / Cmax_fig10 - 1),
    rel_to_isar0 = Cmax / Cmax[1],
    rel_to_isar0_fig10 = Cmax_fig10 / Cmax_fig10[1]
  )
fig10_cmax |>
  dplyr::select(
    SCORE_ISAR, Cmax, Cmax_fig10, pct_diff, rel_to_isar0, rel_to_isar0_fig10,
    C12, C12_fig10
  ) |>
  dplyr::rename(
    "ISAR" = SCORE_ISAR,
    "Simulated Cmax (ng/mL)" = Cmax,
    "Figure 10 Cmax (ng/mL)" = Cmax_fig10,
    "Cmax difference (%)" = pct_diff,
    "Simulated Cmax, relative to ISAR 0" = rel_to_isar0,
    "Figure 10 Cmax, relative to ISAR 0" = rel_to_isar0_fig10,
    "Simulated C at 12 h (ng/mL)" = C12,
    "Figure 10 C at 12 h (ng/mL)" = C12_fig10
  ) |>
  knitr::kable(digits = 2)

# Deterministic typical-value solve against digitised figure values: the
# absolute peak within the reading error of the figure, and the ISAR-driven
# reduction in the peak (the covariate effect itself) within a few percent.
stopifnot(
  all(abs(fig10_cmax$pct_diff) < 10),
  all(abs(fig10_cmax$rel_to_isar0 - fig10_cmax$rel_to_isar0_fig10) < 0.03)
)
```

The peak and its ISAR dependence are reproduced: the peaks are within
`r round(max(abs(fig10_cmax$pct_diff)))`% of the figure and the ratio to ISAR 0
matches to two decimals, which supports the uncentred covariate form (centring
ISAR on 3 would lower Q at ISAR 0 by a factor of `exp(-0.765) = 0.47` and move
the ISAR 0 peak the other way). The 12 h concentration matches at ISAR 0, but
the figure keeps rising with ISAR at 12 h while the simulation does not. The
figure does not state the CrCL it used, and the late tail depends strongly on
CrCL through `V2/F` and `CL/F`.

# Figure 11 (CrCL)

Figure 11 shows a 50 or 100 mg dose given at 1 h on top of an existing
concentration (about 400, 290 and 200 ng/mL before the 100 mg dose at CrCL 30,
75 and 120 mL/min). The prior regimen that produced those baselines, and the
ISAR used, are not stated. Two things can be checked. First, the rise above
baseline is about 220 ng/mL in all three panels, i.e. the single-dose peak does
not depend on CrCL:

```{r fig11}
fig11_subj <- data.frame(id = 1:3, CRCL = c(30, 75, 120), SCORE_ISAR = 3, dose_mg = 100)
fig11_ev <- dplyr::bind_rows(
  fig11_subj |>
    dplyr::mutate(time = 0, amt = dose_mg, evid = 1L, cmt = "depot", dvid = NA_integer_),
  tidyr::crossing(fig11_subj, time = seq(0, 12, by = 0.05)) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))
fig11_sim <- rxode2::rxSolve(mod_typ, events = fig11_ev, keep = "CRCL", useLinCmt = FALSE) |>
  as.data.frame()
fig11_cmax <- fig11_sim |>
  dplyr::group_by(CRCL) |>
  dplyr::summarise(Cmax = max(Cc), .groups = "drop")
knitr::kable(
  fig11_cmax |> dplyr::rename("CrCL (mL/min)" = CRCL, "Single-dose Cmax (ng/mL)" = Cmax),
  digits = 1
)
stopifnot(all(abs(fig11_cmax$Cmax / 220 - 1) < 0.10))
```

Second, the pre-dose baselines fall about two-fold from CrCL 30 to 120 mL/min.
Typical steady-state troughs after 100 mg every 12 or 24 hours (eight weeks of
dosing, which is several terminal half-lives):

```{r fig11-ss}
make_ss_events <- function(subj, tau, t_end, obs_tad) {
  n_dose <- floor(t_end / tau) + 1
  t_last <- tau * (n_dose - 1)
  dplyr::bind_rows(
    subj |>
      dplyr::mutate(
        time = 0, amt = dose_mg, evid = 1L, cmt = "depot",
        ii = tau, addl = n_dose - 1, dvid = NA_integer_
      ),
    tidyr::crossing(subj, time = t_last + obs_tad) |>
      dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", ii = 0, addl = 0, dvid = 1L)
  ) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    dplyr::mutate(t_last = t_last)
}
t_end <- 24 * 7 * 8
fig11_ss <- dplyr::bind_rows(lapply(c(12, 24), function(tau) {
  ev <- make_ss_events(fig11_subj, tau, t_end, 0)
  rxode2::rxSolve(mod_typ, events = ev, keep = "CRCL", useLinCmt = FALSE, maxsteps = 500000L) |>
    as.data.frame() |>
    dplyr::mutate(interval = paste0("every ", tau, " h"))
})) |>
  dplyr::select(interval, CRCL, trough = Cc) |>
  dplyr::mutate(baseline_fig11 = rep(c(400, 290, 200), 2))
fig11_ss |>
  dplyr::rename(
    "Dosing interval" = interval,
    "CrCL (mL/min)" = CRCL,
    "Simulated steady-state trough (ng/mL)" = trough,
    "Figure 11 baseline (ng/mL)" = baseline_fig11
  ) |>
  knitr::kable(digits = 0)

span <- fig11_ss |>
  dplyr::group_by(interval) |>
  dplyr::summarise(span = trough[CRCL == 30] / trough[CRCL == 120], .groups = "drop")
span

# CrCL acts on CL/F, which is the only tramadol clearance, so the trough must
# fall materially between CrCL 30 and 120 mL/min (Figure 11: about 2-fold). A
# structure in which ODT formation also drains tramadol gives only about 1.2-fold
# here, because CL/F is then a minor share of the total clearance.
stopifnot(all(span$span > 1.4))
```

The baselines span about `r round(max(span$span), 1)`-fold across CrCL 30-120
mL/min in the simulation against about 2-fold in Figure 11. The absolute
baselines depend on the unstated regimen, so they are not asserted.

# Virtual cohort

Covariates are drawn to resemble Table 1. ISAR is an integer 1-5 with
probabilities 0.1, 0.2, 0.4, 0.2, 0.1 (mean 3, SD 1.1). CrCL is log-normal with
median 48 mL/min and log-SD 0.75, truncated to the observed 16-351 mL/min range
(the reported mean 65 and SD 69 indicate a strong right skew). Doses follow the
study mix (25, 50, 75, 100 mg in proportion 3 : 6 : 1 : 11).

```{r cohort}
n_vpc <- 200
draw_crcl <- function(n) {
  x <- stats::rlnorm(n * 3, log(48), 0.75)
  x <- x[x >= 16 & x <= 351]
  x[seq_len(n)]
}
vpc_subj <- data.frame(
  id = seq_len(n_vpc),
  SCORE_ISAR = sample(1:5, n_vpc, replace = TRUE, prob = c(0.1, 0.2, 0.4, 0.2, 0.1)),
  CRCL = draw_crcl(n_vpc),
  dose_mg = sample(c(25, 50, 75, 100), n_vpc, replace = TRUE, prob = c(3, 6, 1, 11) / 21)
)
summary(vpc_subj[, c("SCORE_ISAR", "CRCL", "dose_mg")])
```

# Simulation: steady-state VPC (replicates Figure 6)

The paper's VPC (Figure 6) plots concentrations 0-8 h after a dose in patients
on established therapy. The long-term dosing interval is not reported, so the
same cohort is simulated under both 12-hourly and once-daily dosing for eight
weeks, and the last dosing interval is observed.

```{r vpc-sim}
vpc_sim <- dplyr::bind_rows(lapply(c(12, 24), function(tau) {
  ev <- make_ss_events(vpc_subj, tau, t_end, seq(0, 8, by = 0.5))
  rxode2::rxSolve(
    ui,
    events = ev,
    keep = c("SCORE_ISAR", "CRCL", "dose_mg", "t_last"),
    useLinCmt = FALSE,
    maxsteps = 500000L
  ) |>
    as.data.frame() |>
    dplyr::mutate(interval = paste0("every ", tau, " h"), tad = time - t_last)
}))
stopifnot(!anyNA(vpc_sim$Cc), !anyNA(vpc_sim$Cc_m1))
```

The bands below use individual predictions (`Cc`, `Cc_m1`, between-subject
variability without residual error). Observed medians read off Figure 6 by the
maintainers are overlaid as points.

```{r vpc-plot, fig.height = 6}
vpc_q <- vpc_sim |>
  dplyr::select(interval, id, tad, Tramadol = Cc, ODT = Cc_m1) |>
  tidyr::pivot_longer(c(Tramadol, ODT), names_to = "analyte", values_to = "conc") |>
  dplyr::group_by(interval, analyte, tad) |>
  dplyr::summarise(
    p10 = stats::quantile(conc, 0.1),
    p50 = stats::quantile(conc, 0.5),
    p90 = stats::quantile(conc, 0.9),
    .groups = "drop"
  )
fig6_obs <- data.frame(
  analyte = rep(c("Tramadol", "ODT"), each = 4),
  tad = rep(c(0, 2, 4, 6.8), 2),
  median_obs = c(290, 455, 415, 380, 52, 38, 63, 60)
)
ggplot(vpc_q, aes(tad)) +
  geom_ribbon(aes(ymin = p10, ymax = p90), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = p50), colour = "steelblue") +
  geom_point(data = fig6_obs, aes(y = median_obs), colour = "black") +
  facet_grid(analyte ~ interval, scales = "free_y") +
  labs(
    x = "Time after dose (h)", y = "Concentration (ng/mL)",
    caption = paste(
      "Simulated 10th-90th percentile band and median (individual predictions);",
      "points: observed medians digitised from Figure 6 of Al-Qurain 2022."
    )
  )
```

```{r vpc-table}
vpc_cmp <- vpc_q |>
  dplyr::inner_join(
    fig6_obs |> dplyr::mutate(tad = round(tad * 2) / 2),
    by = c("analyte", "tad")
  ) |>
  dplyr::mutate(ratio_sim_obs = p50 / median_obs)
vpc_cmp |>
  dplyr::select(interval, analyte, tad, p50, median_obs, ratio_sim_obs) |>
  dplyr::arrange(analyte, interval, tad) |>
  dplyr::rename(
    "Dosing interval" = interval,
    "Analyte" = analyte,
    "Time after dose (h)" = tad,
    "Simulated median (ng/mL)" = p50,
    "Figure 6 observed median (ng/mL)" = median_obs,
    "Simulated / observed" = ratio_sim_obs
  ) |>
  knitr::kable(digits = 2)
```

The observed tramadol medians of Figure 6 lie between the two simulated
intervals: once-daily dosing sits close to them and 12-hourly dosing above them.
Under once-daily dosing the simulated ODT medians are about a quarter below the
observed ones at three of the four times and above them at 2 h.
The observed medians come from 99 samples in 21 patients on mixed and unreported
regimens, while the cohort here is drawn from assumed covariate distributions.
The comparison is therefore shown for orientation and not asserted.

## Steady-state ODT/tramadol ratio

Table 1 reports a steady-state trough ODT/tramadol ratio of 0.26 (SD 0.2, range
0.08-0.82; n = 17). The simulated trough ratios (individual predictions at the
first observation of the last dosing interval):

```{r ss-ratio}
ratio_summary <- vpc_sim |>
  dplyr::filter(abs(tad) < 1e-6) |>
  dplyr::mutate(ratio = Cc_m1 / Cc) |>
  dplyr::group_by(interval) |>
  dplyr::summarise(
    mean = mean(ratio),
    sd = stats::sd(ratio),
    median = stats::median(ratio),
    p05 = stats::quantile(ratio, 0.05),
    p95 = stats::quantile(ratio, 0.95),
    .groups = "drop"
  )
knitr::kable(ratio_summary, digits = 3)

# Check on the Kt encoding. With Kt a first-order rate constant on the tramadol
# central amount the typical ratio is Kt*V1/CLm = 0.128, whereas reading the
# printed Kt as a formation clearance would give about 0.34. Assert the cohort
# median, with a band wide enough for any cohort draw that still excludes the
# clearance reading.
stopifnot(
  all(ratio_summary$median > 0.09),
  all(ratio_summary$median < 0.25)
)
```

The median simulated ratio sits near the typical value, and the log-normal
variability on `Kt` and `V1/F` lifts the mean above it toward the observed mean
of 0.26.

# PKNCA validation

The paper reports no non-compartmental results, so the NCA is checked against
the model's own closed forms. After a single oral dose with `F = 1`, the
tramadol `AUC0-inf` is `Dose / CL`. ODT is formed at `Kt x A_central` and
cleared at `CLm / V1`, so its `AUC0-inf` is `Dose x Kt x V1 / (CL x CLm)`. Both
are multiplied by 1000 to give ng*h/mL. Two single-dose arms (50 and 100 mg) of
150 subjects each are simulated without residual error and sampled for six
weeks, because the terminal phase is long.

```{r nca-sim}
n_arm <- 150
nca_subj <- data.frame(
  id = seq_len(2 * n_arm),
  dose_mg = rep(c(50, 100), each = n_arm),
  SCORE_ISAR = sample(1:5, 2 * n_arm, replace = TRUE, prob = c(0.1, 0.2, 0.4, 0.2, 0.1)),
  CRCL = draw_crcl(2 * n_arm)
) |>
  dplyr::mutate(treatment = paste(dose_mg, "mg single dose"))
nca_times <- c(
  0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 12, 16, 24, 36, 48, 72, 96, 120,
  168, 240, 336, 504, 672, 840, 1008
)
nca_ev <- dplyr::bind_rows(
  nca_subj |>
    dplyr::mutate(time = 0, amt = dose_mg, evid = 1L, cmt = "depot", dvid = NA_integer_),
  tidyr::crossing(nca_subj, time = nca_times) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

nca_sim <- rxode2::rxSolve(
  ui |> rxode2::zeroRe("sigma"),
  events = nca_ev,
  keep = c("dose_mg", "treatment"),
  useLinCmt = FALSE,
  # Six weeks out, fast-clearing subjects are many orders of magnitude below
  # their peak; the default tolerances let the solution dip slightly below zero
  # there, which PKNCA rejects.
  atol = 1e-12,
  rtol = 1e-10
) |>
  as.data.frame() |>
  dplyr::mutate(id = as.integer(as.character(id)))
stopifnot(!anyNA(nca_sim$Cc), all(nca_sim$Cc >= 0), all(nca_sim$Cc_m1 >= 0))
```

```{r nca-run}
dose_df <- nca_subj |>
  dplyr::transmute(id, treatment, time = 0, amt = dose_mg)
intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)
run_nca <- function(conc_col) {
  conc_df <- nca_sim |>
    dplyr::transmute(id, treatment, time, conc = .data[[conc_col]]) |>
    dplyr::filter(!is.na(conc))
  conc_obj <- PKNCA::PKNCAconc(conc_df, conc ~ time | treatment + id)
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
  as.data.frame(res$result)
}
nca_parent <- run_nca("Cc") |> dplyr::mutate(analyte = "Tramadol")
nca_m1 <- run_nca("Cc_m1") |> dplyr::mutate(analyte = "ODT")
nca_all <- dplyr::bind_rows(nca_parent, nca_m1)

nca_all |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  dplyr::group_by(analyte, treatment, PPTESTCD) |>
  dplyr::summarise(median = stats::median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  dplyr::rename(
    "Analyte" = analyte,
    "Treatment" = treatment,
    "Cmax (ng/mL)" = cmax,
    "Tmax (h)" = tmax,
    "AUC0-inf (ng*h/mL)" = aucinf.obs,
    "t1/2 (h)" = half.life
  ) |>
  knitr::kable(digits = 2, caption = "Median simulated single-dose NCA by arm.")
```

```{r nca-closed-form}
indiv <- nca_sim |>
  dplyr::group_by(id) |>
  dplyr::slice(1) |>
  dplyr::ungroup() |>
  dplyr::transmute(
    id, dose_mg,
    auc_parent = 1000 * dose_mg / cl,
    auc_m1 = 1000 * dose_mg * kmet * vc / (cl * cl_m1)
  )
chk <- nca_all |>
  dplyr::filter(PPTESTCD == "aucinf.obs") |>
  dplyr::select(id, analyte, auc_nca = PPORRES) |>
  dplyr::left_join(indiv, by = "id") |>
  dplyr::mutate(
    auc_closed = ifelse(analyte == "Tramadol", auc_parent, auc_m1),
    pct_diff = 100 * (auc_nca / auc_closed - 1)
  )
chk |>
  dplyr::group_by(analyte) |>
  dplyr::summarise(
    median_pct_diff = stats::median(pct_diff, na.rm = TRUE),
    p90_abs_pct_diff = stats::quantile(abs(pct_diff), 0.9, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::rename(
    "Analyte" = analyte,
    "Median difference, NCA vs closed form (%)" = median_pct_diff,
    "90th percentile of absolute difference (%)" = p90_abs_pct_diff
  ) |>
  knitr::kable(digits = 2)

# Trapezoidal error and the log-linear tail extrapolation depend on how far
# each subject's terminal phase has progressed by the last sample, which varies
# with the CrCL drawn. Assert the centre and a robust quantile, not the extreme;
# a mis-scaled volume or clearance, or a formation term that also drained
# tramadol, moves these by tens of percent.
for (a in c("Tramadol", "ODT")) {
  d <- chk$pct_diff[chk$analyte == a]
  stopifnot(
    abs(stats::median(d, na.rm = TRUE)) < 5,
    stats::quantile(abs(d), 0.9, na.rm = TRUE) < 15
  )
}
```

# Assumptions and deviations

* **Units of Table 3 volumes and clearances.** Table 3 labels `V1/F`, `V2/F`,
  `Q`, `CL/F` and `CLm/F` in litres and L/h, but the printed values (for example
  `V1/F` 0.373, `CL/F` 0.00604) are about 1000-fold smaller than any plausible
  tramadol value in litres. Monolix applies no unit conversion, so with dose in
  mg and concentration in ng/mL the fitted volume unit is mg/(ng/mL) = 1000 L.
  Every volume and clearance is therefore encoded as the printed value x 1000.
  The Figure 10 peak (above) confirms the scaling.
* **`Kt` is a first-order rate constant (1/h), not a clearance.** The Methods
  call it a clearance and Table 3 prints it in L/h, but the Table 3 footnote and
  Results 3.3 call it a first-order rate constant, and the Figure 6/7 ODT/tramadol
  medians (about 0.13-0.18) agree with `Kt V1 / CLm` = 0.128 rather than with the
  0.34 a clearance reading gives. The L/h label is treated as a misprint and
  the value is used unscaled.
* **ODT formation does not deplete tramadol.** The paper prints no equations.
  Methods 2.4.2 describes tramadol as 'either eliminated from the system with a
  clearance CL or transformed ... into a metabolite', which on its own suggests
  two parallel loss routes, but the same section states that the fraction
  metabolised is not identifiable and only `Vm/(F km)` and `CLm/(F km)` can be
  determined. The model therefore treats `CL/F` as the total tramadol clearance
  and the ODT pool as an apparent amount scaled by `1/Fm`. The maintainers chose
  this reading because the alternative, in which `Kt` also removes tramadol,
  makes `Kt x V1` (18.4 L/h) the dominant tramadol clearance. Under that
  alternative the steady-state VPC reaches only about half of the Figure 6
  medians, and CrCL (which acts only on `CL/F`) moves the Figure 11 baselines
  about 1.2-fold instead of about 2-fold. A consequence of the chosen reading is
  that `Kt x V1` exceeding `CL/F` is not a mass-balance violation, and the ODT
  amount is not a physical mass.
* **ODT volume.** No ODT volume is reported (Table 3, Figure 1 and the
  one-compartment column of Table S3 all lack one). Following the Monolix
  parent-metabolite parameterisation the authors cite (Bertrand et al.), the ODT
  compartment shares the tramadol central volume `V1/F`; the metabolite volume
  is not identifiable from parent-only oral dosing (Methods 2.4.2). This fixes
  the ODT half-life at `0.693 V1 / CLm` = 1.8 h; steady-state ODT levels do
  not depend on this choice.
* **Covariate equations.** None is printed. The Monolix default for a
  continuous covariate, `log(theta_i) = log(theta_pop) + beta * COV`, uncentred,
  is used for ISAR on `Q` and CrCL on `CL/F` and `V2/F`. Figure 10 supports the
  uncentred form for ISAR; no figure discriminates for CrCL. The Table 3 values
  of `Q`, `CL/F` and `V2/F` are therefore the values at ISAR 0 and CrCL 0, not at
  the typical patient.
* **CrCL** is the raw Cockcroft-Gault value in mL/min, not BSA-normalised.
* **No molecular-weight correction** on the tramadol-to-ODT formation
  (tramadol 263.4 g/mol, ODT 249.3 g/mol); the paper describes none, so the
  difference is absorbed in the apparent ODT parameters.
* **ODT residual error.** The paper names a 'combined' model without the form;
  the Monolix default combined form `y = f + (a + b f) e` (`combined1()`) is
  used. The '(%)' labels on the residual rows of Table 3 are misprints: `b` is
  a proportional fraction and the ODT `a` is in ng/mL.
* **Bioavailability** is fixed to 1 (Methods 2.4.2), so all volumes and
  clearances are apparent (/F).
* **Simulation design.** The long-term dosing interval behind Figure 6 and the
  steady-state ratio is not reported, so 12-hourly and once-daily dosing of the
  study dose mix are both shown and neither is asserted against Figure 6. ISAR
  and CrCL distributions for the virtual cohort are assumptions matched to the
  Table 1 summaries. The CrCL used in Figure 10, and the regimen and ISAR behind
  the Figure 11 baselines, are not stated.
* Figure 6, Figure 10 and Figure 11 values used for comparison were digitised
  by the maintainers from the published figures and carry reading error of a
  few ng/mL.
