---
title: "Vancomycin (Alqurain 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vancomycin (Alqurain 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 6)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r model-load}
ui <- rxode2::rxode(readModelDb("Alqurain_2024_vancomycin"))
```

# Model and source

* Citation: `r ui$reference`
* Description: `r ui$description`
* Article: <https://doi.org/10.2147/DDDT.S496512>
* Supplement (Supplementary Tables 1-9, Supplementary Figure 1):
  <https://www.dovepress.com/get_supplementary_file.php?f=496512.docx>

Alqurain 2024 is a multicentre retrospective population-PK analysis of routine
vancomycin therapeutic-drug-monitoring records from three hospitals in the
Eastern Province of Saudi Arabia. Because the dataset consists exclusively of
**trough** concentrations drawn as part of routine TDM, the authors could not
resolve a distribution phase; a one-compartment model with linear elimination
and IV input was retained on AIC grounds over two- and three-compartment,
lag-time, and Michaelis-Menten alternatives (Supplementary Tables 1-6).

# Population

```{r population}
pop <- ui$population
tibble::tibble(Field = names(pop), Value = vapply(pop, paste, character(1), collapse = "; ")) |>
  knitr::kable(caption = "Study population (Alqurain 2024 Table 1).")
```

172 vancomycin trough concentrations from 124 patients aged 40 years and older
(median 79) admitted to general medical wards between 1 January and
31 December 2022. Patients in the emergency department, intensive care unit or
surgical wards, those with chronic kidney disease, and pregnant women were
excluded, so this is explicitly a **non-critical-care** cohort. Renal function
was nonetheless substantially impaired in most patients: mean Cockcroft-Gault
CrCl 61.1 +/- 48.2 mL/min, with 35% at or below 30 mL/min. Mean body weight was
72 +/- 20 kg and 41% of the cohort were women. Intravenous doses ranged from
500 to 1750 mg, with 1000 mg the most common (40%). Concentrations were assayed
by enzyme-multiplied immunoassay (EMIT) over a 2-50 mg/L calibration range.

The same information is available programmatically via
`readModelDb("Alqurain_2024_vancomycin")()$population`.

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Alqurain_2024_vancomycin.R` carries an in-file
comment naming its source location. They are collected here for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lvc` (V) | 193.65 L (RSE 7.41%) | Table 2, "Final Model / Population" column, row `V (L)` |
| `lcl` (CL intercept) | 3.52 L/h (RSE 62.5%) | Table 2, "Final Model / Population" column, row `CL (L/h)` |
| `e_crp_cl` | -0.05 per mg/L (RSE 28%) | Table 2, "Final Model / Population" column, row `CRP effect on CL` |
| `e_crcl_cl` | 0.0088 per mL/min (RSE 48.4%) | Table 2, "Final Model / Population" column, row `CrCl effect on CL` |
| `etalvc` | variance 0.1089 (= 0.33^2) | Table 2, row `Omega V` = 0.33; Table 2 footnote: "Omega, Inter-individual variability presented as standard deviation" |
| `etalcl` | variance 0.6561 (= 0.81^2) | Table 2, row `Omega Cl` = 0.81; same footnote. Cross-checks against the Results text "reduction in BSV for CL from 173% to 81%" (= 100 x omega) |
| `propSd` | 0.30 | Table 2, "Final Model / Population" column, row `Residual error, b` |
| One compartment, linear elimination, IV input, no lag | n/a | Results, "Population Pharmacokinetic Parameters of Vancomycin"; model-selection detail in Supplementary Tables 1-6 |
| Log-normal individual parameters (`exp(eta)`) | n/a | Methods, "Population Pharmacokinetics Model Development", equation 1 |
| Proportional residual error | n/a | Results: "A proportional error model with coefficient b = 0.4 was used to describe residual variability" |
| Uncentered, untransformed covariate model on CL | n/a | Confirmed against Supplementary Table 9 and Figure 6 -- see the next section |

# The covariate model on clearance

The paper names the two retained covariates and tabulates their coefficients,
but never prints the covariate equation. Three candidate forms are consistent
with a Monolix log-normal parameter and the reported coefficients:

$$\textrm{(A) uncentered linear:}\quad CL_i = 3.52 \cdot e^{-0.05\,\textrm{CRP} + 0.0088\,\textrm{CrCl}} \cdot e^{\eta}$$
$$\textrm{(B) mean-centered:}\quad CL_i = 3.52 \cdot e^{-0.05(\textrm{CRP}-80.6) + 0.0088(\textrm{CrCl}-61.1)} \cdot e^{\eta}$$
$$\textrm{(C) log-transformed:}\quad CL_i = 3.52 \cdot e^{-0.05\ln\textrm{CRP} + 0.0088\ln\textrm{CrCl}} \cdot e^{\eta}$$

**Supplementary Table 9 settles it.** That table reports the median simulated
clearance for each of the nine Monte-Carlo scenario groups, and those medians
span roughly 6000-fold (3 L/h down to 0.0005 L/h). Only form (A) can generate a
span that large from the reported coefficients: form (C) varies CL by about 10%
across the whole CRP range, and form (B) has the right span but is displaced by
orders of magnitude in absolute value.

```{r covariate-form-discrimination}
supp9 <- tibble::tribble(
  ~group, ~CRP, ~CRCL, ~V_supp9, ~CL_supp9,
  "Group 1",   20,  90,  198, 3.0000,
  "Group 2",   85,  90,  195, 0.2000,
  "Group 3",  180,  90,  196, 0.0010,
  "Group 4",   20,  60,  195, 2.2000,
  "Group 5",   70,  60,  195, 0.2000,
  "Group 6",  150,  60,  193, 0.0007,
  "Group 7",   10,  30,  192, 1.6000,
  "Group 8",   70,  30,  195, 0.1000,
  "Group 9",  150,  30,  194, 0.0005
)

supp9 |>
  mutate(
    `A: uncentered` = 3.52 * exp(-0.05 * CRP + 0.0088 * CRCL),
    `B: mean-centered` = 3.52 * exp(-0.05 * (CRP - 80.6) + 0.0088 * (CRCL - 61.1)),
    `C: log-transformed` = 3.52 * exp(-0.05 * log(CRP) + 0.0088 * log(CRCL))
  ) |>
  select(group, CRP, CRCL, `Supp. Table 9` = CL_supp9,
         `A: uncentered`, `B: mean-centered`, `C: log-transformed`) |>
  knitr::kable(
    digits = 4,
    caption = paste(
      "Median simulated clearance (L/h) reported in Alqurain 2024 Supplementary Table 9",
      "versus the three candidate covariate forms. Form A tracks the reported values across",
      "their full ~6000-fold range; form B is displaced by orders of magnitude and form C is",
      "essentially flat."
    )
  )
```

Form (A) is therefore what the packaged model encodes. A direct consequence
worth stating plainly: because the covariate model is **uncentered**,
`exp(lcl) = 3.52 L/h` is the clearance of a hypothetical patient with
`CRP = 0` and `CrCl = 0`, **not** the clearance of a typical patient. At the
cohort mean covariate values (CRP 80.6, CrCl 61.1) the model gives a typical
clearance near 0.11 L/h. See the Assumptions and deviations section.

# Virtual cohort

Original patient-level data are not publicly available. Two cohorts are built
below, each mirroring one published artefact.

```{r cohort}
set.seed(20241220)

n_per_group <- 200L   # cap is 200 participants per arm

# --- Cohort 1: deterministic evaluation at Supplementary Table 9's own
# covariate values (one subject per group, IIV zeroed via eta columns).
scen_subj <- supp9 |>
  mutate(id = row_number(), etalcl = 0, etalvc = 0) |>
  select(id, group, CRP, CRCL, etalcl, etalvc)

scen_events <- bind_rows(
  mutate(scen_subj, time = 0, amt = 1000, evid = 1L, cmt = "central"),
  mutate(scen_subj, time = 1, amt = NA_real_, evid = 0L, cmt = "central")
) |>
  arrange(id, time, desc(evid)) |>
  as.data.frame()

# --- Cohort 2: Figure 6 replication. Nine groups defined by the design
# covariate values printed in the Figure 6 panel labels, 1 g q12h, 48 h.
fig6_design <- tidyr::expand_grid(
  CRCL = c(90, 60, 30),
  CRP  = c(20, 80, 180)
) |>
  mutate(
    group = sprintf("Group %d", row_number()),
    panel = sprintf("%s\nCRP = %g, CrCl = %g", group, CRP, CRCL)
  )

make_cohort <- function(CRP, CRCL, group, panel, n, id_offset) {
  subj <- tibble::tibble(
    id = id_offset + seq_len(n),
    CRP = CRP, CRCL = CRCL, group = group, panel = panel
  )
  doses <- subj |>
    tidyr::crossing(time = seq(0, 36, by = 12)) |>
    mutate(amt = 1000, evid = 1L, cmt = "central")
  obs <- subj |>
    tidyr::crossing(time = seq(0, 48, by = 0.5)) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}

events <- Map(
  make_cohort,
  fig6_design$CRP, fig6_design$CRCL, fig6_design$group, fig6_design$panel,
  MoreArgs = list(n = n_per_group),
  id_offset = (seq_len(nrow(fig6_design)) - 1L) * n_per_group
) |>
  bind_rows() |>
  as.data.frame()

# Disjoint-ID guard (duplicate ids across arms silently merge into one subject).
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
stopifnot(length(unique(events$id)) == n_per_group * nrow(fig6_design))
```

# Simulation

```{r simulate}
mod <- readModelDb("Alqurain_2024_vancomycin")

# Deterministic solve: etas supplied as columns and omega = NA, so no IIV is
# sampled. This reads the model file's own encoding of CL and V back out.
sim_scen <- rxode2::rxSolve(
  mod, events = scen_events, omega = NA,
  keep = c("group", "CRP", "CRCL")
) |>
  as.data.frame()

# Stochastic solve for the Figure 6 replication (IIV sampled per subject).
sim <- rxode2::rxSolve(mod, events = events, keep = c("group", "panel")) |>
  as.data.frame()

stopifnot(nrow(sim) > 0, !all(is.na(sim$Cc)))
```

## Deterministic check against Supplementary Table 9

```{r supp9-check}
supp9_cmp <- sim_scen |>
  filter(time == 1) |>
  select(group, CL_model = cl, V_model = vc) |>
  left_join(select(supp9, group, CRP, CRCL, CL_supp9, V_supp9), by = "group") |>
  mutate(
    `CL % diff` = 100 * (CL_model - CL_supp9) / CL_supp9,
    `V % diff`  = 100 * (V_model  - V_supp9)  / V_supp9
  ) |>
  select(group, CRP, CRCL,
         `V, Supp. T9` = V_supp9, `V, model` = V_model, `V % diff`,
         `CL, Supp. T9` = CL_supp9, `CL, model` = CL_model, `CL % diff`)

supp9_cmp |>
  knitr::kable(
    digits = c(0, 0, 0, 0, 1, 1, 4, 4, 1),
    caption = paste(
      "Packaged model evaluated at the covariate values Supplementary Table 9 itself lists,",
      "versus the median simulated V and CL reported there."
    )
  )
```

```{r supp9-assert}
# V carries no covariates, so the model must land on every tabulated median.
stopifnot(max(abs(supp9_cmp[["V % diff"]])) < 5)
```

Volume of distribution agrees within `r sprintf("%.1f%%", max(abs(supp9_cmp[["V % diff"]])))`
of every tabulated median, which is the expected agreement for the median of a
log-normal with no covariates. For clearance, four groups (1, 3, 4 and 5) land
within 10% of the tabulated median, three (2, 7 and 8) within a factor of about
1.8, and two (6 and 9) differ by roughly a factor of 4-5. That residual
disagreement is a property of the published
table rather than of the encoding: Supplementary Table 9's covariate columns
carry the *median sampled* CRP and CrCl of each simulated group (they differ
from the design values printed on the Figure 6 panels -- 85 vs 80, 10 vs 20,
150 vs 180), and because clearance is exponential in CRP the median of the
group is not the model evaluated at the group's median covariate. The
order-of-magnitude agreement across a 6000-fold range is the discriminating
result; see the covariate-form table above.

# Replicate published figures

```{r figure-6, fig.height = 7}
# Replicates Figure 6 of Alqurain 2024: simulated vancomycin concentrations,
# 1 g IV every 12 h for 48 h, across nine CRP x CrCl scenario groups.
sim |>
  filter(!is.na(Cc)) |>
  group_by(panel, time) |>
  summarise(
    Q10 = quantile(Cc, 0.10),
    Q50 = quantile(Cc, 0.50),
    Q90 = quantile(Cc, 0.90),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), fill = "steelblue", alpha = 0.3) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~panel, ncol = 3) +
  coord_cartesian(ylim = c(0, 30)) +
  scale_x_continuous(breaks = seq(0, 45, by = 15)) +
  labs(
    x = "Time (h)", y = "Vancomycin concentration (mg/L)",
    title = "Figure 6 - simulated trough concentrations by CRP and CrCl group",
    caption = paste(
      "Replicates Figure 6 of Alqurain 2024. Line = median, band = 10th-90th percentile",
      "across 200 simulated subjects per group. Dose 1 g IV q12h."
    )
  ) +
  theme_bw() +
  theme(strip.text = element_text(size = 7))
```

The replication reproduces the three features that carry the paper's argument:
every panel starts at `r sprintf("%.1f", 1000 / 193.65)` mg/L immediately after
the first 1 g dose (= 1000 mg / 193.65 L), concentrations rise with successive
doses, and the high-CRP panels plateau near
`r sprintf("%.1f", 4 * 1000 / 193.65)` mg/L because clearance at CRP = 180 is
effectively zero over a 48-hour window, so the four doses simply accumulate in
the volume of distribution. The low-CRP panels (Groups 1, 4 and 7) peak lower
and turn over after the last dose, which is exactly the CRP-driven separation
the paper reports.

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, group)

# Guarantee a time-zero row per (id, group).
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, group) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, group, time, .keep_all = TRUE) |>
  dplyr::arrange(id, group, time)

stopifnot(nrow(sim_nca) > 0, all(sim_nca$Cc >= 0))

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | group + id)

dose_df <- events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, group)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | group + id)

# Last dosing interval (36-48 h): cmin is the 48-hour trough, which is the
# quantity Figure 6 and the paper's whole argument are about.
intervals <- data.frame(
  start = 36, end = 48,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_summary <- as.data.frame(nca_res) |>
  dplyr::group_by(group, PPTESTCD) |>
  dplyr::summarise(PPORRES = median(PPORRES, na.rm = TRUE), .groups = "drop")

nca_summary |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::rename(
    "Group" = group,
    "Cmax (mg/L)" = cmax,
    "Tmax (h)" = tmax,
    "Ctrough at 48 h (mg/L)" = cmin,
    "AUC36-48 (mg*h/L)" = auclast
  ) |>
  knitr::kable(
    digits = 2,
    caption = "Median PKNCA parameters over the last dosing interval (36-48 h), by scenario group."
  )
```

## Comparison against published values

The source paper reports no NCA parameter table -- with trough-only TDM data
there is no concentration-time profile to integrate. The available published
comparator for exposure is the trough level plotted at the right-hand edge of
each Figure 6 panel, read off the figure below with an estimated
uncertainty of about +/- 1 mg/L.

```{r nca_compare}
# Trough concentrations digitised from the median line at the right edge (~48 h)
# of each Figure 6 panel of Alqurain 2024. Figure-derived, +/- ~1 mg/L.
published <- tibble::tribble(
  ~group,     ~cmin,
  "Group 1",  11.0,
  "Group 2",  18.5,
  "Group 3",  19.5,
  "Group 4",  12.7,
  "Group 5",  18.5,
  "Group 6",  19.0,
  "Group 7",  13.6,
  "Group 8",  19.5,
  "Group 9",  19.7
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_summary,
  reference     = published,
  by            = "group",
  units         = c(cmin = "mg/L"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated 48-hour trough versus the value digitised from the corresponding",
    "Figure 6 panel of Alqurain 2024. * differs from reference by >20%."
  ),
  align = c("l", "l", "r", "r", "r")
)
attr(cmp, "footnote")

# Regression guard: every group must stay inside the figure-read tolerance.
# Observed maximum at the recorded seed is about 11%.
cmp_pct <- as.numeric(gsub("[*%+]", "", cmp[["% diff"]]))
stopifnot(length(cmp_pct) == 9L, !anyNA(cmp_pct), max(abs(cmp_pct)) < 15)
```

Every group agrees with the digitised published trough within the 20%
tolerance (largest deviation
`r sprintf("%.1f%%", max(abs(as.numeric(gsub("[*%+]", "", cmp[["% diff"]])))))`),
which -- given that the reference values are themselves figure reads with
roughly +/- 1 mg/L uncertainty -- is as tight a comparison as this source
supports. The simulated troughs run slightly high across the board, consistent
with reading the median line a little low off a printed figure.

## Closed-form check on the ODE encoding

An independent check that does not depend on any published figure: for a
one-compartment model with instantaneous IV input and linear elimination, the
concentration after $n$ doses of size $D$ given every $\tau$ hours is the
analytic superposition
$C(t) = \frac{D}{V}\sum_{j} e^{-k_{el}(t - j\tau)}$. Comparing that against the
solver output validates the `d/dt(central)` encoding directly.

```{r closed-form}
analytic_trough <- function(CRP, CRCL, dose = 1000, tau = 12, n_dose = 4, t_eval = 48) {
  cl <- 3.52 * exp(-0.05 * CRP + 0.0088 * CRCL)
  vc <- 193.65
  kel <- cl / vc
  dose / vc * sum(exp(-kel * (t_eval - tau * seq(0, n_dose - 1))))
}

typical_events <- events |>
  dplyr::distinct(id, group, CRP, CRCL) |>
  dplyr::group_by(group) |>
  dplyr::slice(1) |>
  dplyr::ungroup() |>
  dplyr::mutate(etalcl = 0, etalvc = 0) |>
  dplyr::left_join(dplyr::select(events, id, time, amt, evid, cmt), by = "id") |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

sim_typical <- rxode2::rxSolve(
  mod, events = typical_events, omega = NA, keep = c("group", "CRP", "CRCL")
) |>
  as.data.frame()

cf <- sim_typical |>
  dplyr::filter(time == 48) |>
  dplyr::mutate(
    analytic = mapply(analytic_trough, CRP, CRCL),
    `% diff` = 100 * (Cc - analytic) / analytic
  ) |>
  dplyr::select(Group = group, CRP, CRCL,
                `Solver Cc at 48 h` = Cc, `Analytic C(48 h)` = analytic, `% diff`)

knitr::kable(
  cf, digits = c(0, 0, 0, 3, 3, 4),
  caption = "rxode2 solver output versus the analytic one-compartment superposition, typical subject."
)

stopifnot(max(abs(cf[["% diff"]])) < 0.1)
```

The solver and the closed form agree to better than 0.1% in every scenario, so
the ODE system, the covariate model, and the unit conventions (mg dosed into
litres giving mg/L) are all encoded as intended.

# Assumptions and deviations

- **The covariate model on CL is uncentered, and this makes `exp(lcl)` an
  extrapolated intercept rather than a typical value.** `exp(lcl) = 3.52 L/h`
  is the clearance at `CRP = 0` and `CrCl = 0`. Evaluated at the cohort mean
  covariates (CRP 80.6 mg/L, CrCl 61.1 mL/min) the model returns a typical
  clearance of about 0.11 L/h and a terminal half-life over 1000 h, which is
  far outside the accepted vancomycin range (roughly 3-6 L/h and 6-12 h in
  adults with preserved renal function). This is a property of the published
  model, not of the transcription: the uncentered form is the only one that
  reproduces Supplementary Table 9 and Figure 6, both of which the authors
  generated from their own fitted model in Simulx. Users should treat this
  model as a faithful record of the publication and should **not** use it for
  dose prediction outside the CRP and CrCl ranges the paper simulated.
- **Table 2 versus the Results prose.** The Results text states "The estimated
  V was 224.4 L and the estimated CL was 3.52 L/h", which pairs the *base*
  model's V (224.37 L) with the *final* model's CL. Table 2's final column
  gives V = 193.65 L, and Supplementary Table 9's median simulated V of
  192-198 L across all nine groups independently confirms 193.65 L as the
  final-model value. The packaged model uses 193.65 L.
- **Table 2 versus Supplementary Table 1 for the base model.** Table 2's base
  column (V 224.37, CL 0.0035, Omega V 0.37, Omega CL 1.73, b 0.38) does not
  match Supplementary Table 1's one-compartment infusion / linear-elimination
  entry (V 180.23, CL 0.94, Omega V 0.34, Omega CL 0.64, b 0.39), even though
  the two agree on the base-model objective function (377.5 / 377.54). Only
  the final model is packaged, so this discrepancy does not affect the model
  file; it is recorded here because it bears on how much weight to place on
  the base-model column.
- **CRP units.** Table 1 and the Figure 6 caption label CRP as `mg/dL`, but the
  reported values (mean 80.6, simulation scenarios 20 / 80 / 180) are two
  orders of magnitude above any plausible mg/dL CRP and match the range
  routinely reported in mg/L for a serious-infection cohort. The canonical
  register unit `mg/L` is used and the source numbers are carried across
  unchanged, so `e_crp_cl` applies per mg/L. The same magnitude concern
  applies to the serum creatinine in Table 1 (183 +/- 171, labelled mg/dL,
  consistent with umol/L), but serum creatinine is not a covariate in the
  final model.
- **CrCl is raw Cockcroft-Gault, not BSA-normalized.** Stored under the
  canonical `CRCL` column, which explicitly accepts raw mL/min with the assay
  form documented per model (precedents in the register:
  `Georges_2009_ceftazidime.R`, `Chen_2023_nemonoxacin.R`,
  `Delattre_2010_amikacin.R`).
- **CRP is a concurrent, not a baseline, covariate.** The authors state the CRP
  values "were measured at the time of vancomycin trough level collection, and
  they are not reflective to the actual inflammatory state when initiating
  vancomycin therapy". The model file records this in
  `covariateData[["CRP"]]$notes`. The simulations here hold CRP constant per
  subject, matching the paper's own Simulx scenarios.
- **Infusion duration is not identifiable and is not part of the model.** The
  paper selected "infusion administration", but Supplementary Table 1 shows the
  bolus and infusion variants returning byte-identical objective functions and
  parameter estimates, so the trough-only data carry no information about
  infusion duration. The simulations here use instantaneous IV input, which is
  what reproduces Figure 6 (every panel is already at 1000/193.65 = 5.2 mg/L at
  t = 0). Users simulating a clinically realistic 1-2 h infusion should supply
  `rate` or `dur` in the event table; over a 12 h interval this changes the
  trough by well under 1%.
- **Simulation cohort covariates are fixed per group, not sampled.** The paper
  stratified 1000 virtual patients into nine groups "based on their levels of
  CRP ... and CrCl", so covariates are held at the scenario values and all
  spread comes from IIV. The Figure 6 replication uses the design values
  printed on the panel labels (CRP 20 / 80 / 180, CrCl 90 / 60 / 30); the
  Supplementary Table 9 check uses that table's own covariate values, which
  differ slightly and appear to be the median *sampled* covariate per group.
- **Non-paper-derived values used for validation only.** The nine trough
  concentrations in the "Comparison against published values" table were read
  off the Figure 6 panels of Alqurain 2024 by on-screen digitisation
  (+/- ~1 mg/L). They are validation targets only -- no model parameter is
  derived from a figure. Every `ini()` value comes from Table 2.
- **Residual error is reported once with two values.** The Results text says
  "A proportional error model with coefficient b = 0.4", while Table 2 gives
  b = 0.38 (base) and 0.3 (final). The final-model value 0.3 is used; 0.4
  appears to be a rounding of the base-model 0.38.
