---
title: "Tramadol and O-desmethyltramadol in older patients (Al-Qurain 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tramadol and O-desmethyltramadol in older patients (Al-Qurain 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
rxode2::rxSetSeed(20220215)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("AlQurain_2022_tramadol"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
cat("* Article: <https://doi.org/10.1007/s13318-022-00756-x>\n", sep = "")
```

Al-Qurain et al. fitted tramadol and its active CYP2D6-formed metabolite
O-desmethyltramadol (ODT) jointly in 21 older hospital inpatients taking
extended-release oral tramadol, using Monolix 2020 (SAEM). The final estimates
are in Table 3 of the paper. The open-access supplement holds the
structural-model and covariate-screening tables (Tables S1-S7) but no model code
and no covariate equation, so several encoding choices below rest on the
reasoning laid out in [Assumptions and deviations](#assumptions-and-deviations).

A search on 2026-10-02 (Europe PMC, and the Crossref record for the DOI) found
no erratum or correction notice for this article.

# Population

`r ui$population$n_subjects` patients aged 69-93 years (mean 83.2, SD 7.5) were
recruited at the Royal Adelaide Hospital (Australia) between June and November
2019 (Table 1). 43% were male; body weight was 33-112 kg (mean 68.3). All had
normal liver function tests. Renal function was wide-ranging: Cockcroft-Gault
creatinine clearance (CrCL) 16-351 mL/min (mean 65.1, SD 69.1). The
Identification of Seniors At Risk (ISAR) frailty score was 1-5 (mean 3, SD 1.2),
and 38% took a CYP3A4 inhibitor and 38% a CYP2D6 inhibitor. Doses were 25 mg
(n = 3), 50 mg (n = 6), 75 mg (n = 1) or 100 mg (n = 11). Up to five
finger-prick volumetric absorptive microsamples were taken per patient within
12 h of a dose (99 samples in total, all above the 10 ng/mL limit of
quantitation).

The same information is available programmatically via
`readModelDb("AlQurain_2022_tramadol")()$population`.

# Model structure

Figure 1 of the paper: oral tramadol is absorbed first-order (`ka`) into a
central compartment (`V1/F`) that exchanges with a peripheral compartment
(`V2/F`, `Q`). Tramadol leaves the central compartment by linear elimination
(`CL/F`) and by first-order conversion (`Kt`) into a one-compartment ODT pool
cleared by `CLm/F`. The ODT pool shares the tramadol central volume (see
Assumptions). Covariates enter as uncentred exponentials:

* `Q = 42.6 * exp(0.255 * ISAR)` L/h
* `CL/F = 6.04 * exp(0.00498 * CrCL)` L/h
* `V2/F = 379 * exp(0.0119 * CrCL)` L

```{r model-states}
# The model defines a cl / vc pair, which rxode2 could replace with an analytic
# linCmt() solution, discarding the metabolite compartment. Confirm the four ODE
# states are present.
stopifnot(identical(
  rxode2::rxState(rxode2::rxModelVars(ui)),
  c("depot", "central", "peripheral1", "central_m1")
))
```

# Source trace

| Quantity | Value in model | Source |
|---|---|---|
| `ka` | 2.96 1/h | Table 3, final model, 'K a, /h' |
| `V1/F` (`vc`) | 373 L | Table 3, 'V 1 / F , l' = 0.373 (x 1000; see Assumptions) |
| `Q` | 42.6 L/h at ISAR 0 | Table 3, 'Q , l/h' = 0.0426 (x 1000) |
| ISAR effect on Q | 0.255 | Table 3, 'ISAR effect on Q' |
| `V2/F` (`vp`) | 379 L at CrCL 0 | Table 3, 'V 2 / F , l' = 0.379 (x 1000) |
| CrCL effect on V2/F | 0.0119 per mL/min | Table 3, 'CrCL effect on V 2 / F' |
| `CL/F` (`cl`) | 6.04 L/h at CrCL 0 | Table 3, 'CL/ F , l/h' = 0.00604 (x 1000) |
| CrCL effect on CL/F | 0.00498 per mL/min | Table 3, 'CrCL effect on CL/F' |
| `Kt` (`kmet`) | 0.0492 1/h | Table 3, 'K t , l/h' (rate constant per footnote) |
| `CLm/F` (`cl_m1`) | 143 L/h | Table 3, 'CL m / F , l/h' = 0.143 (x 1000) |
| IIV (SD) ka, V1, Q, V2, CL, Kt, CLm | 0.378, 0.738, 0.0219, 0.974, 0.0661, 0.785, 0.216 | Table 3 'Omega' rows; footnote: presented as SD |
| Tramadol proportional error | 0.157 | Table 3, 'Tramadol, b (%)' |
| ODT additive error | 5.19 ng/mL | Table 3, 'O -desmethyltramadol, a (%)' |
| ODT proportional error | 0.189 | Table 3, 'O -desmethyltramadol, b (%)' |
| Structure (2-cmt parent, 1-cmt metabolite, first-order absorption and conversion) | `model()` ODEs | Figure 1; Methods 2.4.2; Results 3.3 |
| Log-normal IIV | `exp(theta + eta)` | Methods 2.4.1, Eq. 1 |
| Covariate form | uncentred `exp(beta * COV)` | Not printed; Monolix default, checked against Figure 10 |
| Error models | proportional (tramadol), combined (ODT) | Results 3.3 |

# Typical-value checks

The covariate model is algebraic, so the typical individual parameters at any
covariate value follow exactly from `ini()`. At the cohort means (ISAR 3,
CrCL 65.1 mL/min):

```{r typical-params}
mod_typ <- ui |> rxode2::zeroRe()
ini_vals <- ui$theta
typ <- data.frame(
  SCORE_ISAR = c(0, 3, 6, 3),
  CRCL = c(0, 65.1, 65.1, 120)
) |>
  dplyr::mutate(
    Q = exp(ini_vals[["lq"]] + ini_vals[["e_score_isar_q"]] * SCORE_ISAR),
    CL = exp(ini_vals[["lcl"]] + ini_vals[["e_crcl_cl"]] * CRCL),
    V2 = exp(ini_vals[["lvp"]] + ini_vals[["e_crcl_vp"]] * CRCL),
    CL_form_ODT = exp(ini_vals[["lkmet"]]) * exp(ini_vals[["lvc"]]),
    fm = CL_form_ODT / (CL + CL_form_ODT)
  )
typ |>
  dplyr::rename(
    "ISAR" = SCORE_ISAR,
    "CrCL (mL/min)" = CRCL,
    "Q (L/h)" = Q,
    "CL/F (L/h)" = CL,
    "V2/F (L)" = V2,
    "Kt x V1 (L/h)" = CL_form_ODT,
    "Fraction to ODT" = fm
  ) |>
  knitr::kable(digits = 3)

# The covariate terms reproduce the cohort-mean values quoted in the
# covariateData notes (CL/F 8.3 L/h and V2/F 820 L at CrCL 65.1 mL/min).
stopifnot(
  abs(typ$CL[2] - 8.35) < 0.05,
  abs(typ$V2[2] / 822 - 1) < 0.01
)
```

The typical ODT/tramadol steady-state concentration ratio is `Kt x V1 / CLm`
because the two species share the same volume:

```{r typical-ratio}
ratio_typ <- exp(ini_vals[["lkmet"]]) * exp(ini_vals[["lvc"]]) / exp(ini_vals[["lcl_m1"]])
ratio_typ
```

# Replicate Figure 10 (typical profiles by ISAR)

Figure 10 shows typical single-dose tramadol profiles at ISAR 0, 3 and 6 after
50 and 100 mg, with the dose given at 1 h on the figure axis. The CrCL used is
not stated; the cohort mean 65.1 mL/min is used here (the peak is insensitive
to CrCL, see Figure 11 below).

```{r fig10}
fig10_subj <- tidyr::crossing(
  SCORE_ISAR = c(0, 3, 6),
  dose_mg = c(50, 100)
) |>
  dplyr::mutate(id = dplyr::row_number(), CRCL = 65.1)

fig10_ev <- dplyr::bind_rows(
  fig10_subj |>
    dplyr::mutate(time = 1, amt = dose_mg, evid = 1L, cmt = "depot", dvid = NA_integer_),
  tidyr::crossing(fig10_subj, time = seq(0, 12, by = 0.05)) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

fig10_sim <- rxode2::rxSolve(
  mod_typ,
  events = fig10_ev,
  keep = c("SCORE_ISAR", "dose_mg"),
  useLinCmt = FALSE
) |>
  as.data.frame()

ggplot(fig10_sim, aes(time, Cc, colour = factor(dose_mg))) +
  geom_line() +
  facet_wrap(~ paste("ISAR =", SCORE_ISAR)) +
  scale_colour_manual(values = c("50" = "black", "100" = "red")) +
  labs(
    x = "Time (h)", y = "Tramadol concentration (ng/mL)", colour = "Dose (mg)",
    caption = "Replicates Figure 10 of Al-Qurain 2022."
  )
```

Peak concentrations after 100 mg, against values read off Figure 10 by the
maintainers:

```{r fig10-table}
fig10_cmax <- fig10_sim |>
  dplyr::filter(dose_mg == 100) |>
  dplyr::group_by(SCORE_ISAR) |>
  dplyr::summarise(Cmax = max(Cc), .groups = "drop") |>
  dplyr::mutate(
    Cmax_fig10 = c(240, 220, 195),
    pct_diff = 100 * (Cmax / Cmax_fig10 - 1),
    rel_to_isar0 = Cmax / Cmax[1],
    rel_to_isar0_fig10 = Cmax_fig10 / Cmax_fig10[1]
  )
fig10_cmax |>
  dplyr::rename(
    "ISAR" = SCORE_ISAR,
    "Simulated Cmax (ng/mL)" = Cmax,
    "Figure 10 Cmax (ng/mL)" = Cmax_fig10,
    "Difference (%)" = pct_diff,
    "Simulated, relative to ISAR 0" = rel_to_isar0,
    "Figure 10, relative to ISAR 0" = rel_to_isar0_fig10
  ) |>
  knitr::kable(digits = 2)

# Deterministic typical-value solve against digitised figure values: the
# absolute peak within the reading error of the figure, and the ISAR-driven
# reduction in the peak (the covariate effect itself) within a few percent.
stopifnot(
  all(abs(fig10_cmax$pct_diff) < 10),
  all(abs(fig10_cmax$rel_to_isar0 - fig10_cmax$rel_to_isar0_fig10) < 0.03)
)
```

The peak and its ISAR dependence are reproduced: with Q at ISAR 0 the peak is
about 6% below the figure, and the ratio to ISAR 0 matches to two decimals,
which supports the uncentred covariate form (centring ISAR on 3 would lower Q at
ISAR 0 by a factor of `exp(-0.765) = 0.47` and move the ISAR 0 peak the other
way). The 11 h concentrations of the figure (about 90, 112 and 115 ng/mL) sit
above the simulated values (`r paste(round(fig10_sim$Cc[fig10_sim$dose_mg == 100 & abs(fig10_sim$time - 12) < 1e-6]), collapse = ", ")`
ng/mL), i.e. the published simulation has a slower post-peak decline than the
Table 3 typical values give at CrCL 65.1 mL/min; the figure does not state the
CrCL it used, and the late tail depends strongly on CrCL through `V2/F`.

# Figure 11 (CrCL)

Figure 11 shows a 50 or 100 mg dose given at 1 h on top of an existing
concentration (about 400, 290 and 200 ng/mL before the 100 mg dose at CrCL 30,
75 and 120 mL/min); the prior regimen that produced those baselines, and the
ISAR used, are not stated, so the baselines cannot be reproduced. What can be
checked is the dose increment: the rise above baseline is about 220 ng/mL in
all three panels, i.e. the single-dose peak does not depend on CrCL.

```{r fig11}
fig11_subj <- data.frame(id = 1:3, CRCL = c(30, 75, 120), SCORE_ISAR = 3, dose_mg = 100)
fig11_ev <- dplyr::bind_rows(
  fig11_subj |>
    dplyr::mutate(time = 0, amt = dose_mg, evid = 1L, cmt = "depot", dvid = NA_integer_),
  tidyr::crossing(fig11_subj, time = seq(0, 12, by = 0.05)) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))
fig11_sim <- rxode2::rxSolve(mod_typ, events = fig11_ev, keep = "CRCL", useLinCmt = FALSE) |>
  as.data.frame()
fig11_cmax <- fig11_sim |>
  dplyr::group_by(CRCL) |>
  dplyr::summarise(Cmax = max(Cc), C12 = Cc[abs(time - 12) < 1e-6], .groups = "drop")
fig11_cmax |>
  dplyr::rename(
    "CrCL (mL/min)" = CRCL,
    "Single-dose Cmax (ng/mL)" = Cmax,
    "Concentration at 12 h (ng/mL)" = C12
  ) |>
  knitr::kable(digits = 1)
stopifnot(all(abs(fig11_cmax$Cmax / 220 - 1) < 0.10))
```

The single-dose peak is close to the ~220 ng/mL increment of Figure 11 at every
CrCL, and higher CrCL lowers the concentration later in the interval, which is
the direction of the figure.

# Virtual cohort

Covariates are drawn to resemble Table 1. ISAR is an integer 1-5 with
probabilities 0.1, 0.2, 0.4, 0.2, 0.1 (mean 3, SD 1.1). CrCL is log-normal with
median 48 mL/min and log-SD 0.75, truncated to the observed 16-351 mL/min range
(the reported mean 65 and SD 69 indicate a strong right skew). For the
steady-state cohort the doses follow the study mix (25, 50, 75, 100 mg in
proportion 3 : 6 : 1 : 11), given every 12 hours.

```{r cohort}
n_vpc <- 200
draw_crcl <- function(n) {
  x <- stats::rlnorm(n * 3, log(48), 0.75)
  x <- x[x >= 16 & x <= 351]
  x[seq_len(n)]
}
vpc_subj <- data.frame(
  id = seq_len(n_vpc),
  SCORE_ISAR = sample(1:5, n_vpc, replace = TRUE, prob = c(0.1, 0.2, 0.4, 0.2, 0.1)),
  CRCL = draw_crcl(n_vpc),
  dose_mg = sample(c(25, 50, 75, 100), n_vpc, replace = TRUE, prob = c(3, 6, 1, 11) / 21),
  treatment = "Study dose mix, every 12 h"
)
summary(vpc_subj[, c("SCORE_ISAR", "CRCL", "dose_mg")])
```

# Simulation: steady-state VPC (replicates Figure 6)

The paper's VPC (Figure 6) plots concentrations 0-8 h after a dose in patients
on established therapy. Thirty 12-hourly doses are simulated, and the last
dosing interval is observed.

```{r vpc-sim}
tau <- 12
n_dose <- 30
t_last <- tau * (n_dose - 1)
vpc_ev <- dplyr::bind_rows(
  vpc_subj |>
    dplyr::mutate(
      time = 0, amt = dose_mg, evid = 1L, cmt = "depot",
      ii = tau, addl = n_dose - 1, dvid = NA_integer_
    ),
  tidyr::crossing(vpc_subj, time = t_last + seq(0, 8, by = 0.5)) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", ii = 0, addl = 0, dvid = 1L)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

vpc_sim <- rxode2::rxSolve(
  ui,
  events = vpc_ev,
  keep = c("SCORE_ISAR", "CRCL", "dose_mg"),
  useLinCmt = FALSE
) |>
  as.data.frame() |>
  dplyr::mutate(tad = time - t_last)
stopifnot(!anyNA(vpc_sim$Cc), !anyNA(vpc_sim$Cc_m1))
```

The band below uses individual predictions (`Cc`, `Cc_m1`, between-subject
variability without residual error). Observed medians read off Figure 6 by the
maintainers are overlaid as points.

```{r vpc-plot}
vpc_q <- vpc_sim |>
  dplyr::select(id, tad, Tramadol = Cc, ODT = Cc_m1) |>
  tidyr::pivot_longer(c(Tramadol, ODT), names_to = "analyte", values_to = "conc") |>
  dplyr::group_by(analyte, tad) |>
  dplyr::summarise(
    p10 = stats::quantile(conc, 0.1),
    p50 = stats::quantile(conc, 0.5),
    p90 = stats::quantile(conc, 0.9),
    .groups = "drop"
  )
fig6_obs <- data.frame(
  analyte = rep(c("Tramadol", "ODT"), each = 4),
  tad = rep(c(0, 2, 4, 6.8), 2),
  median_obs = c(290, 455, 415, 380, 52, 38, 63, 60)
)
ggplot(vpc_q, aes(tad)) +
  geom_ribbon(aes(ymin = p10, ymax = p90), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = p50), colour = "steelblue") +
  geom_point(data = fig6_obs, aes(y = median_obs), colour = "black") +
  facet_wrap(~analyte, scales = "free_y") +
  labs(
    x = "Time after dose (h)", y = "Concentration (ng/mL)",
    caption = paste(
      "Simulated 10th-90th percentile band and median (individual predictions);",
      "points: observed medians digitised from Figure 6 of Al-Qurain 2022."
    )
  )
```

```{r vpc-table}
vpc_cmp <- vpc_q |>
  dplyr::inner_join(
    fig6_obs |> dplyr::mutate(tad = round(tad * 2) / 2),
    by = c("analyte", "tad")
  ) |>
  dplyr::mutate(ratio_sim_obs = p50 / median_obs)
vpc_cmp |>
  dplyr::select(analyte, tad, p50, median_obs, ratio_sim_obs) |>
  dplyr::rename(
    "Analyte" = analyte,
    "Time after dose (h)" = tad,
    "Simulated median (ng/mL)" = p50,
    "Figure 6 observed median (ng/mL)" = median_obs,
    "Simulated / observed" = ratio_sim_obs
  ) |>
  knitr::kable(digits = 2)
```

The simulated medians are lower than the observed medians of Figure 6
throughout the interval, more so for tramadol than for ODT. The sampled patients
are not a random draw of the covariate distribution assumed here: the observed
VPC median is built from 99 samples in 21 patients, and the dosing interval of
the long-term regimen is not reported (12-hourly is assumed). The comparison is
therefore shown for orientation rather than asserted.

## Steady-state ODT/tramadol ratio

Table 1 reports a steady-state trough ODT/tramadol ratio of 0.26 (SD 0.2, range
0.08-0.82; n = 17). The simulated trough ratios (individual predictions at the
end of the 29th interval, i.e. the first observation of the last interval):

```{r ss-ratio}
trough <- vpc_sim |>
  dplyr::filter(abs(tad) < 1e-6) |>
  dplyr::mutate(ratio = Cc_m1 / Cc)
ratio_summary <- trough |>
  dplyr::summarise(
    mean = mean(ratio),
    sd = stats::sd(ratio),
    median = stats::median(ratio),
    p05 = stats::quantile(ratio, 0.05),
    p95 = stats::quantile(ratio, 0.95)
  )
knitr::kable(ratio_summary, digits = 3)

# Structural check on the Kt encoding. With Kt a rate constant out of V1 the
# typical ratio is near Kt*V1/CLm = 0.128 (slightly above it at a trough,
# because tramadol is still declining while ODT follows it), whereas reading Kt
# as a clearance would give about 0.34. Assert the cohort median, with a band
# wide enough for any cohort draw that still excludes the clearance reading.
stopifnot(
  ratio_summary$median > 0.09,
  ratio_summary$median < 0.25
)
```

The median simulated ratio sits at the typical value, and the log-normal
variability on `Kt` and `V1/F` lifts the mean above it, toward the observed mean
of 0.26.

# PKNCA validation

The paper reports no non-compartmental results, so the NCA is checked against
the model's own closed forms. After a single oral dose with `F = 1`, the
tramadol `AUC0-inf` is `Dose / CL` and the ODT `AUC0-inf` is
`Dose x fm / CLm`, with `fm = Kt V1 / (CL + Kt V1)`, both multiplied by 1000 to
give ng*h/mL. Two single-dose arms (50 and 100 mg) of 150 subjects each are
simulated without residual error and sampled densely to 168 h, because large
`V2/F` at high CrCL gives a long terminal phase.

```{r nca-sim}
n_arm <- 150
nca_subj <- data.frame(
  id = seq_len(2 * n_arm),
  dose_mg = rep(c(50, 100), each = n_arm),
  SCORE_ISAR = sample(1:5, 2 * n_arm, replace = TRUE, prob = c(0.1, 0.2, 0.4, 0.2, 0.1)),
  CRCL = draw_crcl(2 * n_arm)
) |>
  dplyr::mutate(treatment = paste(dose_mg, "mg single dose"))
nca_times <- c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 12, 16, 24, 36, 48, 72, 96, 120, 144, 168)
nca_ev <- dplyr::bind_rows(
  nca_subj |>
    dplyr::mutate(time = 0, amt = dose_mg, evid = 1L, cmt = "depot", dvid = NA_integer_),
  tidyr::crossing(nca_subj, time = nca_times) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

nca_sim <- rxode2::rxSolve(
  ui |> rxode2::zeroRe("sigma"),
  events = nca_ev,
  keep = c("dose_mg", "treatment"),
  useLinCmt = FALSE
) |>
  as.data.frame() |>
  dplyr::mutate(id = as.integer(as.character(id)))
stopifnot(!anyNA(nca_sim$Cc))
```

```{r nca-run}
dose_df <- nca_subj |>
  dplyr::transmute(id, treatment, time = 0, amt = dose_mg)
intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)
run_nca <- function(conc_col) {
  conc_df <- nca_sim |>
    dplyr::transmute(id, treatment, time, conc = .data[[conc_col]]) |>
    dplyr::filter(!is.na(conc))
  conc_obj <- PKNCA::PKNCAconc(conc_df, conc ~ time | treatment + id)
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
  as.data.frame(res$result)
}
nca_parent <- run_nca("Cc") |> dplyr::mutate(analyte = "Tramadol")
nca_m1 <- run_nca("Cc_m1") |> dplyr::mutate(analyte = "ODT")
nca_all <- dplyr::bind_rows(nca_parent, nca_m1)

nca_all |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  dplyr::group_by(analyte, treatment, PPTESTCD) |>
  dplyr::summarise(median = stats::median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  dplyr::rename(
    "Analyte" = analyte,
    "Treatment" = treatment,
    "Cmax (ng/mL)" = cmax,
    "Tmax (h)" = tmax,
    "AUC0-inf (ng*h/mL)" = aucinf.obs,
    "t1/2 (h)" = half.life
  ) |>
  knitr::kable(digits = 2, caption = "Median simulated single-dose NCA by arm.")
```

```{r nca-closed-form}
indiv <- nca_sim |>
  dplyr::group_by(id) |>
  dplyr::slice(1) |>
  dplyr::ungroup() |>
  dplyr::transmute(
    id, dose_mg,
    auc_parent = 1000 * dose_mg / cl,
    auc_m1 = 1000 * dose_mg * (kmet * vc / (cl + kmet * vc)) / cl_m1
  )
chk <- nca_all |>
  dplyr::filter(PPTESTCD == "aucinf.obs") |>
  dplyr::select(id, analyte, auc_nca = PPORRES) |>
  dplyr::left_join(indiv, by = "id") |>
  dplyr::mutate(
    auc_closed = ifelse(analyte == "Tramadol", auc_parent, auc_m1),
    pct_diff = 100 * (auc_nca / auc_closed - 1)
  )
chk |>
  dplyr::group_by(analyte) |>
  dplyr::summarise(
    median_pct_diff = stats::median(pct_diff, na.rm = TRUE),
    p90_abs_pct_diff = stats::quantile(abs(pct_diff), 0.9, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::rename(
    "Analyte" = analyte,
    "Median difference, NCA vs closed form (%)" = median_pct_diff,
    "90th percentile of absolute difference (%)" = p90_abs_pct_diff
  ) |>
  knitr::kable(digits = 2)

# Trapezoidal error and the log-linear tail extrapolation depend on how far
# each subject's terminal phase has progressed by 168 h, which varies with the
# CrCL drawn. Assert the centre and a robust quantile, not the extreme; a
# mis-scaled volume or clearance, or a wrong conversion route, moves these by
# tens of percent.
for (a in c("Tramadol", "ODT")) {
  d <- chk$pct_diff[chk$analyte == a]
  stopifnot(
    abs(stats::median(d, na.rm = TRUE)) < 5,
    stats::quantile(abs(d), 0.9, na.rm = TRUE) < 15
  )
}
```

# Assumptions and deviations

* **Units of Table 3 volumes and clearances.** Table 3 labels `V1/F`, `V2/F`,
  `Q`, `CL/F` and `CLm/F` in litres and L/h, but the printed values (for example
  `V1/F` 0.373, `CL/F` 0.00604) are about 1000-fold smaller than any plausible
  tramadol value in litres. Monolix applies no unit conversion, so with dose in
  mg and concentration in ng/mL the fitted volume unit is mg/(ng/mL) = 1000 L.
  Every volume and clearance is therefore encoded as the printed value x 1000.
  The Figure 10 peak (above) confirms the scaling.
* **`Kt` is a first-order rate constant (1/h), not a clearance.** The Methods
  call it a clearance and Table 3 prints it in L/h, but the Table 3 footnote and
  Results 3.3 call it a first-order rate constant, and the Figure 6/7 ODT/tramadol
  medians (about 0.13-0.15) agree with `Kt V1 / CLm` = 0.128 rather than with the
  0.34 a clearance reading gives. The L/h label is treated as a misprint and
  the value is used unscaled.
* **ODT volume.** No ODT volume is reported (Table 3, Figure 1 and the
  one-compartment column of Table S3 all lack one). Following the Monolix
  parent-metabolite parameterisation the authors cite (Bertrand et al.), the ODT
  compartment shares the tramadol central volume `V1/F`; the metabolite volume
  is not identifiable from parent-only oral dosing (Methods 2.4.2). This fixes
  the ODT half-life at `0.693 V1 / CLm` = 1.8 h; steady-state ODT levels do
  not depend on this choice.
* **Covariate equations.** None is printed. The Monolix default for a
  continuous covariate, `log(theta_i) = log(theta_pop) + beta * COV`, uncentred,
  is used for ISAR on `Q` and CrCL on `CL/F` and `V2/F`. Figure 10 supports the
  uncentred form for ISAR; no figure discriminates for CrCL. The Table 3 values
  of `Q`, `CL/F` and `V2/F` are therefore the values at ISAR 0 and CrCL 0, not at
  the typical patient.
* **CrCL** is the raw Cockcroft-Gault value in mL/min, not BSA-normalised.
* **No molecular-weight correction** on the tramadol-to-ODT transfer
  (tramadol 263.4 g/mol, ODT 249.3 g/mol); the paper describes none, so the
  difference is absorbed in the apparent `CLm/F`.
* **ODT residual error.** The paper names a 'combined' model without the form;
  the Monolix default combined form `y = f + (a + b f) e` (`combined1()`) is
  used. The '(%)' labels on the residual rows of Table 3 are misprints: `b` is
  a proportional fraction and the ODT `a` is in ng/mL.
* **Bioavailability** is fixed to 1 (Methods 2.4.2), so all volumes and
  clearances are apparent (/F).
* **Simulation design.** The long-term dosing interval behind Figure 6 and the
  steady-state ratio is not reported; 12-hourly dosing of the study dose mix is
  assumed. ISAR and CrCL distributions for the virtual cohort are assumptions
  matched to the Table 1 summaries. The CrCL used in Figure 10, and the regimen
  and ISAR behind the Figure 11 baselines, are not stated.
* Figure 6, Figure 10 and Figure 11 values used for comparison were digitised
  by the maintainers from the published figures and carry reading error of a
  few ng/mL.
