---
title: "Batefenterol PK and trough FEV1 dose-response in COPD (Ambery 2015)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Batefenterol PK and trough FEV1 dose-response in COPD (Ambery 2015)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7.5, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Ambery 2015 reports two models that were fitted independently and are
**not** linked through exposure: a plasma population PK model and an
empirical Emax dose-response model for day-29 trough FEV1 driven by the
total daily dose. They are packaged as two model files sharing this
vignette.

* PK citation: `r rxode2::rxode(readModelDb("Ambery_2015_batefenterol"))$reference`
* PK description: `r rxode2::rxode(readModelDb("Ambery_2015_batefenterol"))$description`
* PD description: `r rxode2::rxode(readModelDb("Ambery_2015_batefenterol_fev1"))$description`
* Article (open access): <https://doi.org/10.1007/s40268-015-0104-x>
* Parent trial: GSK MAB115032 / <https://clinicaltrials.gov/study/NCT01319019>

GSK961081 (batefenterol) is a bifunctional molecule carrying a muscarinic
antagonist and a beta2-agonist joined by an inert linker, delivered as an
inhaled dry powder (DISKUS).

# Population

The parent study was a 4-week multicenter, randomized, double-blind,
double-dummy, placebo- and salmeterol-controlled parallel-group trial in
outpatients aged 40 years or above with moderate-to-severe stable COPD
(post-bronchodilator FEV1/FVC below 70% and FEV1 30-70% of the NHANES III
predicted value), who were current or former smokers with at least a
10 pack-year history. Patients on a stable inhaled-corticosteroid dose were
eligible; diagnosed asthma was an exclusion. Three once-daily doses (100,
400 and 800 ug) and three twice-daily doses (100, 200 and 400 ug) of
GSK961081 DISKUS were studied (Ambery 2015 Sects. 2.1 and 2.2).

Ambery 2015 Table 1 describes two overlapping analysis sets. The
**PD / PK-PD set** (n = 347) had a mean age of 63 +/- 8.2 years, weight
76 +/- 14 kg, height 171 +/- 8.5 cm, 65% male, 49% current smokers, 58%
concurrent inhaled-corticosteroid use, and a day-1 baseline trough FEV1 of
1.31 +/- 0.46 L. The **PK model-building set** (n = 47) had a mean age of
63 +/- 8.9 years, weight 72 +/- 13 kg, height 169 +/- 7.7 cm, 62% male,
49% current smokers, 53% concurrent ICS use, and a day-1 baseline trough
FEV1 of 1.36 +/- 0.42 L.

The PK assay had a lower limit of quantification of 25 pg/mL, and more than
50% of the concentrations were below it for every treatment except 800 ug
once daily (days 1 and 28) and 400 ug twice daily (day 28). The PK
model-building data set was therefore restricted to those two arms on day
28 only (405 samples from 47 patients, 30% and 27% below LLOQ
respectively), with the below-LLOQ records handled by the M3 likelihood
method. The full PK data set used for the visual predictive check comprised
1112 samples from 128 patients across all six GSK961081 arms
(Ambery 2015 Sects. 2.4 and 3.1).

The same metadata is available programmatically via
`rxode2::rxode(readModelDb("Ambery_2015_batefenterol"))$population` and
`rxode2::rxode(readModelDb("Ambery_2015_batefenterol_fev1"))$population`.

# Source trace

## Pharmacokinetic model (`Ambery_2015_batefenterol`)

A two-compartment disposition model with first-order absorption, coded in
NONMEM as ADVAN4 TRANS4 (Ambery 2015 Sect. 3.1). Plasma concentrations were
"expressed as natural logarithms" and fitted by Monte Carlo importance
sampling with M3 handling of below-LLOQ data (Sect. 2.4). Table 2 reports
the structural THETAs **on the log scale** (its column header is
"Log estimate"), so the model file transcribes them verbatim rather than
back-transforming and re-logging.

| Equation / parameter | Value | Source location |
|----------------------|------:|-----------------|
| Two-compartment, first-order absorption (ADVAN4 TRANS4) | n/a | Sect. 3.1 |
| `lka` (log KA) | -0.890 | Table 2 (95% CI -1.16 to -0.625); KA = 0.411 /h |
| `lcl` (log CL/F) | 6.85 | Table 2 (95% CI 6.62 to 7.08); CL/F = 945 L/h |
| `lvc` (log V2/F) | 6.26 | Table 2 (95% CI 5.82 to 6.72); V2/F = 523 L |
| `lq` (log Q) | 7.25 | Table 2 (95% CI 6.94 to 7.56); Q = 1408 L/h |
| `lvp` (log V3/F) | 9.97 | Table 2 (95% CI 9.40 to 10.5); V3/F = 21,402 L |
| `etalcl` (omega^2 on log CL/F) | 0.594 | Table 2 row "Interindividual variability CL/F" (SE 0.137, 95% CI 0.325 to 0.863) |
| `expSd` (residual SD on log conc) | 0.6042 = sqrt(0.365) | Table 2 row "Proportional residual error" = 0.365 (SE 0.0175, 95% CI 0.331 to 0.399) |
| No covariates retained | n/a | Sect. 3.1; Discussion (Ribbing and Jonsson, ref. 22) |
| IIV only on CL/F | n/a | Sect. 3.1 |
| LLOQ 25 pg/mL; 30% / 27% BLQ in the two model-building arms | n/a | Sects. 2.3 and 2.4 |

## Pharmacodynamic model (`Ambery_2015_batefenterol_fev1`)

Ambery 2015 Sect. 3.2 gives the model as two displayed equations:

$$
\text{Effect} \;=\; (E_0 \times \text{COV}) \;+\; E_{\max}\,\frac{\text{Total Dose}}{\text{Total Dose} + \mathrm{ED}_{50}}
\qquad\text{(Eq. 1)}
$$

$$
\text{COV} \;=\; \mathrm{CON} \times \frac{\text{Baseline FEV}_1\text{ on Day 1}}{\text{Median Baseline FEV}_1\text{ on Day 1}}
\qquad\text{(Eq. 2)}
$$

"where Emax is the trough FEV1 at the maximum effect, E0 is the FEV1 at
zero dose, ED50 is the dose producing 50% of the maximum effect, and CON is
the baseline FEV1 covariate effect."

| Equation / parameter | Value | Source location |
|----------------------|------:|-----------------|
| Emax structure in total daily dose (Eq. 1) | n/a | Sect. 3.2 |
| Baseline-FEV1 covariate on the intercept (Eq. 2) | n/a | Sect. 3.2 |
| `le0` (intercept for E0) | 0.0650 L | Table 3 (RSE 32.8%, 95% CI 0.0233 to 0.107) |
| `e_fev1_bl_e0` (CON) | 19.0 | Table 3 row "Baseline (L)" (RSE 36.9%, 95% CI 5.24 to 32.8) |
| `lemax` (Emax) | 0.293 L | Table 3 (RSE 14.9%, 95% CI 0.207 to 0.379) |
| `led50` (ED50) | 152 ug/day | Table 3 (RSE 50.2%, 95% CI 2.45 to 302) |
| `propSd_FEV1` (proportional residual SD) | 0.204 | Table 3 (RSE 30.8%, 95% CI 0.0809 to 0.327) |
| Median baseline FEV1 normaliser | **not reported**; 1.31 L (Table 1 mean) substituted | Table 1, PD / PK-PD column |
| Only baseline FEV1 retained from the covariate screen | n/a | Sects. 2.4 and 3.2 |
| Once- and twice-daily arms pooled on total daily dose | n/a | Sect. 4 |
| Day-29 trough FEV1 = mean of 11 h and 12 h post evening dose on day 28 | n/a | Sect. 2.3 |

# Errata

No published erratum, corrigendum or correction notice was located for
Ambery 2015. The PubMed Central record (PMC4561049, PMID 26286203) carries
no "Erratum in", "Correction" or "Retraction" notice, and a literature
search for a correction to `doi:10.1007/s40268-015-0104-x` returned none,
checked on 2026-09-15. The article is open access under a Creative Commons
Attribution-NonCommercial 4.0 licence.

# Virtual cohort

Original individual records are not publicly available. The cohorts below
approximate the published trial design and demographics.

```{r cohort}
# `set.seed()` seeds R's RNG. It does NOT seed rxode2's simulation RNG, whose
# streams are partitioned per solver thread -- so the rxode2-drawn PK cohort
# below is reproducible on this machine and different on a machine with a
# different thread count. Every assertion downstream is written so that it
# holds for ANY cohort the model can produce.
set.seed(20150819)

n_per_arm <- 100L

# ---- PK cohort: the two arms that formed the PK model-building data set,
# dosed for 28 days so that day 28 is at steady state (the terminal
# half-life of the model is about 26 h, so 28 days is many half-lives).
# Day 28 spans 648-672 h; the paper sampled after both the morning and the
# evening dose on that day (Sect. 2.3).
day28_start <- 27 * 24        # 648 h -- the day-28 morning dose
obs_times   <- seq(day28_start, day28_start + 24, by = 0.25)

make_pk_arm <- function(n, amt, ii, addl, arm, id_offset = 0L) {
  ids <- id_offset + seq_len(n)
  doses <- data.frame(
    id = ids, time = 0, amt = amt, evid = 1L,
    ii = ii, addl = addl, cmt = "depot", arm = arm
  )
  obs <- tidyr::crossing(id = ids, time = obs_times) |>
    dplyr::mutate(amt = 0, evid = 0L, ii = 0, addl = 0L,
                  # cmt on an observation row must be a declared ODE state,
                  # never the algebraic observable `Cc`.
                  cmt = "central", arm = arm)
  dplyr::bind_rows(doses, obs) |> dplyr::arrange(id, time, dplyr::desc(evid))
}

pk_events <- dplyr::bind_rows(
  # 800 ug once daily: 28 doses, the last at 648 h.
  make_pk_arm(n_per_arm, amt = 800, ii = 24, addl = 27L,
              arm = "800 ug QD", id_offset = 0L),
  # 400 ug twice daily: 56 doses, the last at 660 h.
  make_pk_arm(n_per_arm, amt = 400, ii = 12, addl = 55L,
              arm = "400 ug BID", id_offset = 1000L),
  # A sub-therapeutic arm carried only to reproduce the paper's statement
  # that every other treatment was more than 50% below LLOQ.
  make_pk_arm(n_per_arm, amt = 100, ii = 24, addl = 27L,
              arm = "100 ug QD", id_offset = 2000L)
)
stopifnot(!anyDuplicated(unique(pk_events[, c("id", "time", "evid")])))

# ---- PD cohort: the five total daily dose levels that appear on the x-axis
# of Ambery 2015 Fig. 3d and Fig. 4. Baseline FEV1 is drawn to match the
# Table 1 PD-set summary (mean 1.31 L, SD 0.46 L) and truncated to the
# plausible spirometric range for the enrolled population; the paper reports
# only the mean and SD, not the shape of the distribution.
pd_doses <- c(0, 100, 200, 400, 800)
pd_events <- lapply(seq_along(pd_doses), function(k) {
  data.frame(
    id   = (k - 1L) * 1000L + seq_len(n_per_arm),
    time = 0, amt = 0, evid = 0L,
    DOSE_BATEFENTEROL_UGD = pd_doses[k],
    FEV1_BL = pmin(pmax(rnorm(n_per_arm, mean = 1.31, sd = 0.46), 0.40), 3.00),
    dose_lbl = pd_doses[k]
  )
}) |> dplyr::bind_rows()
stopifnot(!anyDuplicated(pd_events$id))
```

# Simulation

```{r simulate}
mod_pk     <- readModelDb("Ambery_2015_batefenterol")
mod_pk_typ <- rxode2::zeroRe(mod_pk)

mod_pd     <- readModelDb("Ambery_2015_batefenterol_fev1")
mod_pd_typ <- rxode2::zeroRe(mod_pd)

# `omega` is passed EXPLICITLY on both PK solves. rxode2 keeps `omega` in the
# solve options attached to the COMPILED model, and the two rxUi objects here
# share one compiled ODE system -- so without these arguments the second solve
# silently inherits the first solve's omega. That fails in both directions: the
# typical-value solve would re-sample etas (a random draw masquerading as the
# typical patient), and, had the order been reversed, the population solve would
# collapse every subject onto one patient. Both failures are SILENT. The
# `stopifnot()` guards below are what make them detectable.

# Stochastic PK cohort: `Cc` is the individual prediction and `sim` carries
# the residual error, i.e. `sim` is what the assay would have reported.
sim_pk <- rxode2::rxSolve(mod_pk, events = pk_events, keep = c("arm"),
                          omega = rxode2::rxode(mod_pk)$omega) |>
  as.data.frame()
# IIV must actually have been sampled.
stopifnot(dplyr::n_distinct(round(sim_pk$cl, 8)) > 1L)

# Typical-value PK (no IIV, no residual) for the mass-balance / NCA gate.
sim_pk_typ <- rxode2::rxSolve(mod_pk_typ, events = pk_events, keep = c("arm"),
                              omega = NA) |>
  as.data.frame()
# ...and must NOT have been sampled here: every subject is the typical patient.
stopifnot(dplyr::n_distinct(round(sim_pk_typ$cl, 8)) == 1L,
          abs(unique(round(sim_pk_typ$cl, 8)) - exp(6.85)) < 1e-6)

# Typical-value PD surface. The PD model is algebraic (no ODE state), so a
# single row per subject is the whole simulation. NOTE: `omega = NA` is NOT
# passed here -- this model declares no eta at all, and rxode2 evaluates
# `rep(0, dim(NA)[1])` and errors when the sentinel is given to a model with no
# random effects.
sim_pd_typ <- rxode2::rxSolve(mod_pd_typ, events = pd_events,
                              keep = c("dose_lbl")) |>
  as.data.frame()
```

# Replicate Figure 2 of Ambery 2015 -- PK visual predictive check

Figure 2 plots the natural logarithm of the plasma GSK961081 concentration
in pg/mL against time on day 28, one panel per treatment, with a horizontal
line at the LLOQ (log(25) = `r round(log(25), 2)`) and dashed 95% prediction
intervals. Panels (c) and (f) of the source figure are the two arms that
formed the model-building data set.

```{r figure-2, fig.cap = "Replicates Figure 2 of Ambery 2015 (panels c and f, the two model-building arms, plus the 100 ug once-daily arm of panel d). Solid line: simulated median; ribbon: simulated 95 percent prediction interval; dashed horizontal line: the 25 pg/mL assay LLOQ."}
vpc <- sim_pk |>
  dplyr::filter(!is.na(sim)) |>
  # The model works in ug/L; the paper plots log(pg/mL) and 1 ug/L = 1000 pg/mL.
  dplyr::mutate(log_pg_mL = log(sim * 1000),
                tad = time - day28_start) |>
  dplyr::group_by(arm, tad) |>
  dplyr::summarise(
    Q025 = quantile(log_pg_mL, 0.025),
    Q50  = quantile(log_pg_mL, 0.500),
    Q975 = quantile(log_pg_mL, 0.975),
    .groups = "drop"
  )

ggplot(vpc, aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q025, ymax = Q975), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = log(25), linetype = "dashed") +
  facet_wrap(~arm) +
  labs(x = "Time after the day-28 morning dose (h)",
       y = "log plasma GSK961081 concentration (pg/mL)",
       caption = "Replicates Figure 2 of Ambery 2015.")
```

## The paper's below-LLOQ dichotomy

Sect. 2.4 states that more than 50% of the PK data were below the 25 pg/mL
LLOQ for every treatment **except** 800 ug once daily and 400 ug twice
daily on day 28, where 30% and 27% respectively were below it. That
dichotomy is a direct, non-tuned consequence of the reported clearance and
residual error, so it is the strongest structural check available for the
PK model (the paper publishes no NCA table).

```{r blq-check}
blq <- sim_pk |>
  dplyr::filter(!is.na(sim)) |>
  dplyr::group_by(arm) |>
  dplyr::summarise(pct_blq = 100 * mean(sim * 1000 < 25), .groups = "drop") |>
  dplyr::arrange(pct_blq)

blq |>
  dplyr::rename("Treatment arm" = arm, "Simulated % below LLOQ" = pct_blq) |>
  knitr::kable(digits = 1,
               caption = "Simulated percentage of day-28 concentrations below the 25 pg/mL LLOQ. Ambery 2015 Sect. 2.4 reports 30% for 800 ug once daily and 27% for 400 ug twice daily, and more than 50% for every other treatment.")

pct <- setNames(blq$pct_blq, blq$arm)

# The two model-building arms must land on the reported side of the paper's
# own 50% dichotomy, and the sub-therapeutic arm on the other side. Bounds
# are deliberately wide: the exact percentage depends on the sampling
# schedule, which the paper gives only as five sampling WINDOWS rather than
# nominal times, and on which cohort rxode2 happens to draw (its RNG streams
# are partitioned per solver thread, so CI draws a different cohort).
stopifnot(
  pct[["800 ug QD"]]  < 50,
  pct[["400 ug BID"]] < 50,
  pct[["100 ug QD"]]  > 50
)
```

# PKNCA validation of the PK model

The paper reports no NCA parameters, so the NCA here is a self-consistency
gate on the packaged model rather than a comparison against published
values. At steady state the clearance identity `CL/F * AUCtau = dose per
interval` must hold exactly for the typical-value solve; any error in a
transcribed clearance, volume or dose unit breaks it immediately.

```{r pknca}
# Do NOT add `time > 0` or `Cc > 0` here -- both would drop the interval-start
# record that anchors AUC.
# Every subject in the typical-value solve is identical (no IIV), so one
# representative subject per arm is enough and keeps pk.nca() fast.
nca_conc <- sim_pk_typ |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(arm) |>
  dplyr::filter(id == min(id)) |>
  dplyr::ungroup() |>
  dplyr::select(id, time, Cc, arm)

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ time | arm + id)

nca_dose <- pk_events |>
  dplyr::filter(evid == 1L, id %in% unique(nca_conc$id)) |>
  dplyr::select(id, time, amt, arm) |>
  # Re-state the dose at the start of the observed steady-state interval so
  # PKNCA anchors the interval on a dose record.
  dplyr::mutate(time = day28_start)

dose_obj <- PKNCA::PKNCAdose(nca_dose, amt ~ time | arm + id)

# One full dosing interval on day 28: 24 h for the once-daily arms, 12 h for
# the twice-daily arm.
intervals <- data.frame(
  start   = day28_start,
  end     = c(day28_start + 24, day28_start + 12),
  cmax    = TRUE,
  tmax    = TRUE,
  auclast = TRUE,
  cmin    = TRUE
)
intervals <- dplyr::bind_rows(
  dplyr::mutate(intervals[1, ], arm = "800 ug QD"),
  dplyr::mutate(intervals[2, ], arm = "400 ug BID"),
  dplyr::mutate(intervals[1, ], arm = "100 ug QD")
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "cmin")) |>
  dplyr::group_by(arm, PPTESTCD) |>
  dplyr::summarise(value = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = value)
```

```{r pknca-massbalance}
cl_typ <- exp(6.85)   # CL/F from Table 2, on the linear scale
dose_per_tau <- c("800 ug QD" = 800, "400 ug BID" = 400, "100 ug QD" = 100)

mb <- nca_wide |>
  dplyr::mutate(
    `Dose per interval (ug)`   = unname(dose_per_tau[arm]),
    `CL/F * AUCtau (ug)`       = cl_typ * auclast,
    `Difference (%)`           = 100 * (cl_typ * auclast -
                                          unname(dose_per_tau[arm])) /
                                 unname(dose_per_tau[arm]),
    `Cmax (pg/mL)`             = 1000 * cmax,
    `Ctrough (pg/mL)`          = 1000 * cmin,
    # PKNCA reports tmax relative to the dose, not on the absolute time axis.
    `Tmax (h after dose)`      = tmax
  ) |>
  dplyr::select(
    "Treatment arm"          = arm,
    "Dose per interval (ug)",
    "CL/F * AUCtau (ug)",
    "Difference (%)",
    "Cmax (pg/mL)",
    "Ctrough (pg/mL)",
    "Tmax (h after dose)"
  )

knitr::kable(mb, digits = c(0, 0, 1, 2, 1, 1, 2),
             caption = "Steady-state mass balance on the typical-value solve. At steady state CL/F * AUCtau equals the dose delivered per interval; the residual difference is the linear-trapezoid error of the 0.25 h observation grid.")

# Deterministic quantity (typical-value solve, no cohort draw), so a tight
# bound is correct here. The only error source is trapezoidal integration on
# a 0.25 h grid, which biases AUC slightly low.
stopifnot(max(abs(mb[["Difference (%)"]])) < 2)

# Dose proportionality across the two once-daily arms is exact in a linear
# model; this catches an accidental non-linearity or a mis-scaled dose.
auc_qd <- mb[["CL/F * AUCtau (ug)"]][mb[["Treatment arm"]] == "800 ug QD"] /
  mb[["CL/F * AUCtau (ug)"]][mb[["Treatment arm"]] == "100 ug QD"]
stopifnot(abs(auc_qd - 8) < 0.05)
```

The simulated steady-state average concentration on 800 ug once daily is
`r round(1000 * 800 / exp(6.85) / 24, 1)` pg/mL against a 25 pg/mL LLOQ,
which is what puts that arm -- and only that arm plus 400 ug twice daily --
on the quantifiable side of the paper's below-LLOQ split.

# Replicate Figures 3d and 4 of Ambery 2015 -- trough FEV1 dose-response

## Typical dose-response curve and the paper's own claims

```{r pd-curve, fig.cap = "Day-29 trough FEV1 dose-response for a patient at the cohort-median day-1 baseline FEV1. Points mark the five total daily dose levels studied; the dashed line marks the zero-dose (placebo) prediction."}
dose_grid <- c(0, seq(10, 900, by = 10))
ev_curve <- data.frame(
  id = seq_along(dose_grid), time = 0, amt = 0, evid = 0L,
  DOSE_BATEFENTEROL_UGD = dose_grid,
  FEV1_BL  = 1.31,              # the median-baseline patient: COV normalises to CON
  dose_lbl = dose_grid
)

curve <- rxode2::rxSolve(mod_pd_typ, events = ev_curve, keep = c("dose_lbl")) |>
  as.data.frame() |>
  dplyr::transmute(dose = dose_lbl, fev1 = FEV1)

e0_cov <- curve$fev1[curve$dose == 0]

ggplot(curve, aes(dose, fev1)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = e0_cov, linetype = "dashed") +
  geom_point(data = dplyr::filter(curve, dose %in% pd_doses), size = 2) +
  labs(x = "Total daily dose (ug)", y = "Day 29 trough FEV1 (L)",
       caption = "Replicates the model line of Figure 4a of Ambery 2015 at the median baseline FEV1.")
```

```{r pd-claims}
# Fail loudly rather than returning numeric(0): a lookup that matches no row
# would silently make every assertion below vacuously TRUE.
fev1_at <- function(d) {
  v <- curve$fev1[curve$dose == d]
  if (length(v) != 1L) stop("no unique dose-grid row at ", d, " ug/day")
  v
}
delta <- function(d) fev1_at(d) - e0_cov            # placebo-corrected, L

claims <- tibble::tibble(
  Claim = c(
    "Zero-dose day-29 trough FEV1 at the median baseline equals E0 * CON = 0.0650 * 19.0 L (Table 3)",
    "Increment at the ED50 dose of 152 ug/day is half of Emax (Table 3); the dose grid has no point at 152 ug so the closed form is evaluated",
    "Increment at 400 ug/day lies inside the 155-277 mL trough-FEV1 improvement reported for the parent trial (Sect. 1, ref. 8)",
    "Increment at 800 ug/day lies inside the same 155-277 mL range",
    "'Only a small increase in the trough FEV1 from 400 to 800 ug' (Sect. 4): the 400 -> 800 step is less than half the 0 -> 100 step"
  ),
  Achieved = c(
    sprintf("%.4f L (target %.4f L)", e0_cov, 0.0650 * 19.0),
    sprintf("%.4f L (target %.4f L)", 0.293 * 152 / (152 + 152), 0.293 / 2),
    sprintf("%.0f mL", 1000 * delta(400)),
    sprintf("%.0f mL", 1000 * delta(800)),
    sprintf("%.0f mL vs %.0f mL", 1000 * (delta(800) - delta(400)), 1000 * delta(100))
  )
)
knitr::kable(claims, caption = "Deterministic checks of the packaged PD model against statements and parameter values printed in Ambery 2015.")

# All of these are deterministic (typical-value, single-patient) quantities,
# so tight tolerances are correct.
stopifnot(
  abs(e0_cov - 0.0650 * 19.0) < 1e-6,
  # The dose grid has no point at exactly 152 ug, so evaluate the closed form.
  abs((0.293 * 152 / (152 + 152)) - 0.293 / 2) < 1e-12,
  1000 * delta(400) > 155, 1000 * delta(400) < 277,
  1000 * delta(800) > 155, 1000 * delta(800) < 277,
  (delta(800) - delta(400)) < 0.5 * delta(100)
)
```

## Figure 4a -- day-29 trough FEV1 versus total daily dose

Figure 4a of the source plots the observed day-29 trough FEV1 against total
daily dose together with the model median and its 95% prediction interval.
The prediction interval at a given dose is dominated by the spread of the
day-1 baseline FEV1 covariate (1.31 +/- 0.46 L) acting through Eq. 2, with
the proportional residual error on top.

```{r figure-4a, fig.cap = "Replicates Figure 4a of Ambery 2015. Points: simulated median day-29 trough FEV1 per total daily dose; error bars: simulated 95 percent prediction interval over the virtual cohort's baseline-FEV1 distribution plus proportional residual error."}
# rxSolve returns FEV1 as the individual prediction; the proportional
# residual is applied here in R (rather than via rxode2's `sim` column) so
# that the draw is reproducible from `set.seed()` regardless of thread count.
prop_sd <- 0.204
pd_pi <- sim_pd_typ |>
  dplyr::mutate(fev1_obs = FEV1 * (1 + rnorm(dplyr::n(), 0, prop_sd))) |>
  dplyr::group_by(dose_lbl) |>
  dplyr::summarise(
    Q025 = quantile(fev1_obs, 0.025),
    Q50  = quantile(fev1_obs, 0.500),
    Q975 = quantile(fev1_obs, 0.975),
    .groups = "drop"
  )

ggplot(pd_pi, aes(dose_lbl, Q50)) +
  geom_pointrange(aes(ymin = Q025, ymax = Q975)) +
  geom_line(data = curve, aes(dose, fev1), inherit.aes = FALSE,
            linewidth = 0.8) +
  expand_limits(y = c(0, 3.5)) +
  labs(x = "Total daily dose (ug)", y = "Day 29 trough FEV1 (L)",
       caption = "Replicates Figure 4a of Ambery 2015.")

pd_pi |>
  dplyr::mutate(Width = Q975 - Q025) |>
  dplyr::rename("Total daily dose (ug)" = dose_lbl,
                "2.5th percentile (L)"  = Q025,
                "Median (L)"            = Q50,
                "97.5th percentile (L)" = Q975,
                "95% PI width (L)"      = Width) |>
  knitr::kable(digits = 2,
               caption = "Simulated day-29 trough FEV1 distribution by total daily dose. Ambery 2015 Fig. 4a shows a 95 percent prediction interval of roughly 0.65-2.35 L at zero dose (a width near 1.7 L).")

# The width of the zero-dose prediction interval is the test that decides
# whether Table 3's "Proportional residual error = 0.204" is a standard
# deviation or a variance -- a variance would mean a 45.2% residual CV and
# would widen this interval to roughly 2.9 L, well outside what Fig. 4a
# shows. The bound is wide enough to survive any cohort draw (this is a
# cohort-derived statistic) while still failing the variance reading.
w0 <- pd_pi$Q975[pd_pi$dose_lbl == 0] - pd_pi$Q025[pd_pi$dose_lbl == 0]
stopifnot(w0 > 1.2, w0 < 2.5)
```

## Figure 4b -- placebo-corrected change from baseline

```{r figure-4b, fig.cap = "Replicates the model line of Figure 4b of Ambery 2015: day-29 placebo-corrected change from baseline trough FEV1 versus total daily dose. Points mark the four active total daily dose levels."}
ggplot(dplyr::mutate(curve, delta_mL = 1000 * (fev1 - e0_cov)),
       aes(dose, delta_mL)) +
  geom_line(linewidth = 0.8) +
  geom_point(data = dplyr::filter(
    dplyr::mutate(curve, delta_mL = 1000 * (fev1 - e0_cov)),
    dose %in% pd_doses[pd_doses > 0]), size = 2) +
  labs(x = "Total daily dose (ug)",
       y = "Day 29 placebo-corrected change in trough FEV1 (mL)",
       caption = "Replicates Figure 4b of Ambery 2015.")
```

Reading the model medians off Figure 4b of the source gives roughly 122,
175, 232 and 263 mL at 100, 200, 400 and 800 ug/day; the packaged model
returns `r paste(sprintf("%.0f", 1000 * sapply(pd_doses[pd_doses > 0], function(d) curve$fev1[curve$dose == d] - e0_cov)), collapse = ", ")` mL
at the same doses. The agreement is within the precision with which a
printed figure can be read.

# Assumptions and deviations

- **The median baseline FEV1 that Eq. 2 normalises by is not reported
  anywhere in the paper.** The Table 1 **mean** for the n = 347 PD analysis
  set, 1.31 L, is substituted. Because Eq. 2 is a ratio, the substitution
  cancels exactly for a patient at that value and moves predictions only
  for patients away from the cohort centre; a patient at 2.0 L, for
  instance, is predicted 1.31/median times higher than under the true
  normaliser. Fig. 4a's model line at zero dose (about 1.28 L) is
  consistent with the product `E0 * CON = 1.235 L` that the substitution
  reproduces.

- **Table 2 and Table 3 do not share a convention for their variability
  terms, and each was adjudicated separately against the paper's own
  figures.** Neither table states whether its value is a variance, a
  standard deviation or a coefficient of variation.
    - *Table 2 (PK) reports variances.* In its last two rows the column
      headed "RSE (%)" actually holds the standard error rather than a
      percentage (0.594 +/- 1.96 * 0.137 reproduces the printed
      0.325-0.863 confidence interval; 0.137% does not), which is the
      signature of a block pasted from a NONMEM run summary. The variance
      reading is confirmed by Fig. 2: in panel (f) the 95% prediction
      interval 20 h after the day-28 dose spans roughly 2.3 log units
      either side of the population mean, which needs a total variance of
      order 1. The variance reading gives `0.594 + 0.365` (SD about 0.98);
      the standard-deviation reading gives `0.353 + 0.133` (SD about 0.70),
      which is far too narrow. The packaged model therefore uses
      `etalcl ~ 0.594` and `expSd = sqrt(0.365) = 0.6042`.
    - *Table 3 (PD) reports a standard deviation.* Its RSE column is a
      genuine percentage throughout. A variance reading would imply a 45.2%
      residual CV, which Fig. 3a refutes (the observed day-29 trough FEV1
      scatters about +/-0.4 L around a population prediction of 1.0 L, and
      that spread already contains any inter-individual term) and which
      would also overshoot the 0.65-2.35 L prediction interval that Fig. 4a
      shows at zero dose. The vignette's Figure 4a chunk carries this as a
      live assertion on the simulated interval width.

- **The PD model carries no inter-individual variability, because none is
  reported.** Table 3 lists no IIV term, yet Fig. 3b (observed versus
  individual predictions) is materially tighter than Fig. 3a (observed
  versus population predictions), which can only happen if the fitted model
  contained an eta. With a single trough-FEV1 observation per patient that
  eta would not be separately identifiable from the residual in any case,
  and no variance is published for it, so none has been invented. Simulated
  between-subject spread in the PD figures above therefore comes only from
  the baseline-FEV1 covariate and the residual error, and understates the
  spread of the original fit.

- **The PK model's `Cc` is in ug/L**, which equals ng/mL. Ambery 2015
  tabulates and plots pg/mL; the vignette multiplies by 1000 wherever the
  simulation is compared with the paper. No parameter was rescaled.

- **All PK disposition parameters are apparent (`/F`).** The inhaled
  bioavailability of the DISKUS formulation is not identifiable from
  inhaled-only data and is not reported, so no `f(depot)` term is applied
  and the apparent volumes (V3/F = 21,402 L) are much larger than any
  physiologic volume.

- **The PK cohort is dosed for 28 days to reach the day-28 steady state**
  that the model was fitted to, using `ii` / `addl` rather than `ss = 1`,
  because the model is coded as explicit ODEs. The terminal half-life
  implied by the parameters is about 26 h, so 28 days of dosing is many
  half-lives and the day-28 interval is at steady state.

- **The paper gives PK sampling as five time WINDOWS** (1 h to 0 min
  pre-dose, then 0-30 min, 30 min-2 h, 2-6 h and 6-11 h post-dose) rather
  than nominal times, so the below-LLOQ percentages here are computed over
  a uniform 0.25 h grid across the whole day-28 interval and are not
  expected to reproduce the paper's 30% and 27% exactly. The assertion is
  written against the paper's own 50% dichotomy instead.

- **The virtual baseline-FEV1 distribution is normal, truncated to
  0.40-3.00 L.** Ambery 2015 Table 1 reports only a mean and a standard
  deviation; the truncation keeps draws inside the spirometric range
  implied by the FEV1 30-70%-predicted entry criterion.

- **Neither model was fitted to the placebo or salmeterol arms' PK**
  (those samples were not assayed, Sect. 2.3). Placebo enters the PD model
  only as a total daily dose of 0.

- **The exposure-safety analyses of Sect. 3.3** (maximum glucose, heart
  rate, QTcF and minimum potassium against individual Cmax) are reported as
  graphical explorations with no fitted model and no parameter estimates
  ("Exploratory PK/PD analysis by way of generalized linear models did not
  show any significant relationships (data not shown)"), so there is
  nothing to extract from them.
