---
title: "Sepantronium (Aoyama 2012)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sepantronium (Aoyama 2012)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r meta}
mod_meta <- nlmixr2est::nlmixr(readModelDb("Aoyama_2012_sepantronium"))$meta
```

* Citation: `r mod_meta$reference`
* Description: `r mod_meta$description`
* Article (DOI): <https://doi.org/10.1007/s10637-012-9867-x>

This vignette validates the packaged `Aoyama_2012_sepantronium` model -- a
one-compartment IV population PK model for sepantronium bromide (YM155), a
small-molecule survivin suppressant given as a 7-day continuous IV infusion
(CIVI) every 21 days, in 96 adult patients with non-small-cell lung cancer
(NSCLC), hormone-refractory prostate cancer (HRPC; modern term castration-
resistant prostate cancer / CRPC), or unresectable stage III/IV melanoma --
against the source publication's Table 3 (final-model parameter estimates),
Table 4 (bootstrap confidence intervals), and the paper's explicit covariate-
effect prediction that moderately impaired renal function (CLCR = 40 mL/min)
contributes a 25% decrease in CL and consequently a 34% increase in steady-
state plasma concentration (CSS) during the CIVI.

# Population

The Aoyama 2012 analysis pooled 578 plasma sepantronium concentrations from
96 adult patients (17 female, 79 male) enrolled in three open-label
multicenter Phase 2 trials: LUCY (NSCLC, 33 patients, Europe), PACY (HRPC,
34 patients, North America), and MACY (unresectable stage III/IV melanoma,
29 patients, North America). Median age was 64 years (range 29-90); median
body weight 81 kg (range 50-114); median BSA 1.97 m^2 (range 1.48-2.40);
median Cockcroft-Gault creatinine clearance 79 mL/min (range 31-180); median
ALT 20 U/L (range 6-185). All HRPC patients were male and all NSCLC patients
were Caucasian; ECOG performance status was 0-1 in melanoma and 0-2 in the
other two trials. Each patient received sepantronium 4.8 mg/m^2/day as a
continuous IV infusion over 7 days (168 h) every 21 days; the first dose was
computed from actual body surface area. Concentrations were measured by
LC-MS/MS at PPD Central Laboratory with a lower limit of quantitation of
0.05 ng/mL.

The same information is available programmatically via the model's
`population` metadata:

```{r population}
str(mod_meta$population)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Aoyama_2012_sepantronium.R`.
The table below collects them in one place; values come from Aoyama 2012
Table 3 final-model "Estimate" column unless otherwise noted.

| Parameter / equation                                              | Value      | Source location |
|-------------------------------------------------------------------|------------|-----------------|
| `lcl` (Clearance)                                                 | log(42.1)  | Table 3 row "CL (L/h)" final model |
| `lvc` (Central volume of distribution)                            | log(319)   | Table 3 row "V (L)" final model |
| `e_crcl_cl` (CLCR power exponent on CL)                           | 0.425      | Table 3 row "CL CR (power)" final model |
| `e_alt_cl`  (ALT power exponent on CL)                            | 0.124      | Table 3 row "ALT (power)" final model |
| `e_hrpc_cl` (HRPC proportional effect on CL)                      | -0.045     | Table 3 row "HRPC (ratio)" = 0.955; proportional form (0.955 - 1) |
| `e_mel_cl`  (Melanoma proportional effect on CL)                  | +0.24      | Table 3 row "MM (ratio)" = 1.24; proportional form (1.24 - 1) |
| `etalcl ~ 0.0385`                                                 | 0.0385     | Table 3 row "Random effect on CL" final model |
| `propSd <- sqrt(0.0934)`                                          | 0.306      | Table 3 row "Random effect on Cp (Residual error) w/o outliers" final model; sqrt of reported sigma^2 |
| `cl <- exp(lcl + etalcl) * (CRCL/79)^0.425 * (ALT/20)^0.124 * (1 + e_hrpc_cl * TUMTP_HRPC) * (1 + e_mel_cl * TUMTP_MEL)` | n/a | Methods Eqs. (1), (3), (4) and final-model expression Eq. shown after Table 2 |
| `d/dt(central)` (one-compartment elimination)                     | n/a        | Methods page 444 ("A one-compartment model was applied ... ADVAN1 and TRANS2"); Eq. (1) |
| `Cc ~ prop(propSd)`                                               | n/a        | Methods Eq. (2) proportional residual error |

The paper also reports a second proportional residual-error term
(`sigma^2 = 31.6`) applied to 11 of 96 patients classified as having one or
more "outlier" concentrations (post-hoc IQR rule, `Q3 + 3 * IQR = 23.13
ng/mL`). That second residual error is data-driven and is documented in the
Assumptions and deviations section rather than encoded in the model.

# Virtual cohort

The original observed sepantronium concentrations are not publicly
available. The virtual cohort below approximates the Aoyama 2012 Table 2
demographics: 96 adult patients across the three cancer cohorts (NSCLC
n = 33, HRPC n = 34, melanoma n = 29), with body surface area and the
continuous covariates CRCL and ALT drawn to span the published ranges and
medians. Each subject receives a single first cycle of 4.8 mg/m^2/day CIVI
over 168 h, sampled during the CIVI period and after stop of CIVI, with the
21-day cycle continuing as the observation window.

```{r cohort}
set.seed(20260520)

# Cohort sizes from Aoyama 2012 Table 1
n_nsclc <- 33L
n_hrpc  <- 34L
n_mel   <- 29L
n_total <- n_nsclc + n_hrpc + n_mel

# BSA: log-normal centered on the reported median 1.97 m^2 with SD chosen
# so the simulated range covers ~ Table 2 (1.48-2.40 m^2). Constrain to the
# observed bounds.
bsa_draw <- function(n) {
  s <- exp(rnorm(n, mean = log(1.97), sd = log(2.40 / 1.48) / 4))
  pmin(pmax(s, 1.48), 2.40)
}

# CRCL: log-normal centered on the reported median 79 mL/min, range 31-180.
crcl_draw <- function(n) {
  s <- exp(rnorm(n, mean = log(79), sd = log(180 / 31) / 4))
  pmin(pmax(s, 31), 180)
}

# ALT: log-normal centered on the reported median 20 U/L, range 6-185.
alt_draw <- function(n) {
  s <- exp(rnorm(n, mean = log(20), sd = log(185 / 6) / 4))
  pmin(pmax(s, 6), 185)
}

# Helper to build one cancer-type cohort with disjoint id range.
make_cohort <- function(n, cancer, id_offset = 0L) {
  tibble::tibble(
    id         = id_offset + seq_len(n),
    cancer     = cancer,
    TUMTP_HRPC = as.integer(cancer == "HRPC"),
    TUMTP_MEL  = as.integer(cancer == "melanoma"),
    BSA        = bsa_draw(n),
    CRCL       = crcl_draw(n),
    ALT        = alt_draw(n)
  )
}

cov_tab <- bind_rows(
  make_cohort(n_nsclc, "NSCLC",    id_offset = 0L),
  make_cohort(n_hrpc,  "HRPC",     id_offset = n_nsclc),
  make_cohort(n_mel,   "melanoma", id_offset = n_nsclc + n_hrpc)
)

stopifnot(nrow(cov_tab) == n_total,
          anyDuplicated(cov_tab$id) == 0L)

# Per-cycle dose and CIVI duration. Cycle 1 only -- 7 days CIVI followed by
# 14 days off; observation window 0-21 days.
civi_duration_h <- 7 * 24      # 168 h CIVI
cycle_duration_h <- 21 * 24    # 504 h total cycle
sample_times_h <- c(0, 1, 4, 12, 24, 48, 96, 168,            # during CIVI
                    168.5, 169, 170, 172, 176, 192, 240, 336, 504) # after CIVI

make_subject <- function(idx, row) {
  amt  <- 4.8 * row$BSA * 7            # mg total over the 168-h CIVI
  rate <- amt / civi_duration_h        # mg/h constant rate
  doses <- tibble::tibble(
    id   = idx,           time = 0,
    evid = 1L,            amt  = amt,
    rate = rate,          dv   = NA_real_
  )
  obs <- tibble::tibble(
    id   = idx,           time = sample_times_h,
    evid = 0L,            amt  = NA_real_,
    rate = NA_real_,      dv   = NA_real_
  )
  bind_rows(doses, obs) |>
    mutate(
      cancer     = row$cancer,
      TUMTP_HRPC = row$TUMTP_HRPC,
      TUMTP_MEL  = row$TUMTP_MEL,
      BSA        = row$BSA,
      CRCL       = row$CRCL,
      ALT        = row$ALT
    ) |>
    arrange(time, desc(evid))
}

events <- bind_rows(lapply(seq_len(nrow(cov_tab)), function(i) {
  make_subject(idx = cov_tab$id[i], row = cov_tab[i, ])
}))

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod         <- readModelDb("Aoyama_2012_sepantronium")
mod_typical <- rxode2::zeroRe(mod)

sim_typical <- rxode2::rxSolve(
  object = mod_typical, events = events,
  keep   = c("cancer", "TUMTP_HRPC", "TUMTP_MEL", "BSA", "CRCL", "ALT")
) |>
  as.data.frame()

sim_stoch <- rxode2::rxSolve(
  object = mod, events = events,
  keep   = c("cancer", "TUMTP_HRPC", "TUMTP_MEL", "BSA", "CRCL", "ALT")
) |>
  as.data.frame()
```

# Replicate published figures

## Figure 1 -- plasma sepantronium concentration-time profiles

Aoyama 2012 Figure 1 shows individual plasma sepantronium concentrations
during and after the 7-day CIVI at 4.8 mg/m^2/day. Observed concentrations
during the CIVI cluster around a typical Css of ~5-15 ng/mL with notable
between-subject variability; concentrations decline rapidly after stop of
CIVI consistent with a terminal half-life of ~5 h.

```{r figure-1}
# Replicates Figure 1 of Aoyama 2012: individual concentration-time profiles
# across the 7-day CIVI and 24 h post-CIVI window.
sim_stoch |>
  filter(time > 0, time <= 192) |>
  ggplot(aes(time, Cc, group = id, colour = cancer)) +
  geom_line(alpha = 0.45) +
  geom_vline(xintercept = 168, linetype = "dashed", colour = "gray40") +
  scale_x_continuous(breaks = c(0, 24, 48, 72, 96, 120, 144, 168, 192)) +
  scale_y_continuous(limits = c(0, 40)) +
  labs(
    x = "Time after start of CIVI (h)",
    y = "Plasma sepantronium (ng/mL)",
    colour   = "Cancer type",
    title    = "Figure 1 -- plasma sepantronium concentration-time profiles",
    subtitle = "Replicates Aoyama 2012 Figure 1; 4.8 mg/m^2/day CIVI over 7 days. Dashed line marks end of CIVI."
  ) +
  theme_minimal()
```

## Typical-value steady-state across cancer types

The paper's primary exposure metric is the steady-state plasma sepantronium
concentration during the CIVI (CSS), computed as `CSS = dose_rate / CL`.
Holding all covariates at their reference values, the model predicts a
typical Css around 7-12 ng/mL for a BSA of ~1.97 m^2 with reasonable CRCL
and ALT, and the cancer-type effect should give melanoma > NSCLC > HRPC by
proportional changes of +24% and -4.5% respectively (relative to NSCLC).

```{r typical-css}
typical_css <- sim_typical |>
  filter(time > 96, time <= 168) |>     # late portion of CIVI -> near Css
  group_by(cancer) |>
  summarise(
    median_css = median(Cc, na.rm = TRUE),
    q05        = quantile(Cc, 0.05, na.rm = TRUE),
    q95        = quantile(Cc, 0.95, na.rm = TRUE),
    .groups    = "drop"
  )
knitr::kable(typical_css,
             caption = "Typical-value (zeroRe) steady-state Cc (ng/mL) by cancer type during the last 72 h of CIVI.")
```

## CRCL covariate sensitivity -- the paper's 25% / 34% prediction

Aoyama 2012 explicitly predicts that a patient with moderately impaired
renal function (CLCR = 40 mL/min, vs the population reference of 79 mL/min)
has 25% lower CL and consequently a 34% higher steady-state plasma
sepantronium concentration. The relation `(40/79)^0.425 = 0.749` reproduces
the 25.1% CL reduction; the reciprocal `1 / 0.749 - 1 = 33.5%` reproduces
the 33.5% Css increase. The figure below shows the simulated Css ratio
across the published CRCL range relative to the reference at 79 mL/min for
an otherwise typical NSCLC patient (ALT 20 U/L, BSA 1.97 m^2).

```{r crcl-sensitivity}
crcl_grid <- seq(31, 180, by = 1)
crcl_factor_cl  <- (crcl_grid / 79)^0.425
crcl_factor_css <- 1 / crcl_factor_cl

published_points <- tibble::tibble(
  CRCL = c(40, 79),
  ratio_cl  = (CRCL / 79)^0.425,
  ratio_css = 1 / ratio_cl
)

ggplot(
  tibble::tibble(CRCL = crcl_grid,
                 ratio_cl  = crcl_factor_cl,
                 ratio_css = crcl_factor_css),
  aes(CRCL)
) +
  geom_line(aes(y = ratio_cl,  colour = "CL ratio")) +
  geom_line(aes(y = ratio_css, colour = "Css ratio (= 1 / CL ratio)")) +
  geom_hline(yintercept = 1.0,  linetype = "dotted") +
  geom_vline(xintercept = 79,   linetype = "dotted") +
  geom_vline(xintercept = 40,   linetype = "dashed", colour = "gray40") +
  geom_point(data = published_points,
             aes(y = ratio_cl), colour = "#1f77b4", size = 3) +
  geom_point(data = published_points,
             aes(y = ratio_css), colour = "#d62728", size = 3) +
  scale_colour_manual(values = c("CL ratio" = "#1f77b4",
                                 "Css ratio (= 1 / CL ratio)" = "#d62728")) +
  labs(
    x = "CLCR (mL/min)",
    y = "Ratio (vs typical CLCR = 79 mL/min)",
    colour = NULL,
    title  = "CRCL covariate sensitivity on CL and Css",
    subtitle = "Dashed line at CLCR = 40 mL/min reproduces the paper's 25% CL drop / 34% Css rise."
  ) +
  theme_minimal()
```

```{r crcl-sensitivity-table}
knitr::kable(
  tibble::tibble(
    CLCR_mL_per_min = c(31, 40, 79, 100, 180),
    CL_ratio_vs_typical  = round((CLCR_mL_per_min / 79)^0.425, 3),
    Css_ratio_vs_typical = round(1 / (CLCR_mL_per_min / 79)^0.425, 3)
  ),
  caption = "CRCL covariate effect on CL and Css (typical NSCLC patient)."
)
```

# PKNCA validation

The Aoyama 2012 paper does not tabulate Cmax / Cmin / AUC by patient
stratum; the published exposure metric is Css computed analytically as
`dose / CL` (paper Eq. 6). PKNCA is nonetheless used here to summarise
Cmax, Cmin (at end of CIVI), AUClast (over the full 21-day cycle), and
AUC over the CIVI window per cancer type, as an internal sanity check on
the simulation pipeline. The PKNCA formula carries `cancer` as the
grouping variable so per-cohort comparisons against the model's analytic
Css are possible.

```{r pknca}
sim_for_nca <- sim_stoch |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, cancer) |>
  as.data.frame()

doses_for_nca <- events |>
  filter(evid == 1L) |>
  select(id, time, amt, cancer) |>
  as.data.frame()

conc_obj <- PKNCA::PKNCAconc(
  data    = sim_for_nca,
  formula = Cc ~ time | cancer + id,
  concu   = "ng/mL",
  timeu   = "hr"
)
dose_obj <- PKNCA::PKNCAdose(
  data    = doses_for_nca,
  formula = amt ~ time | cancer + id,
  doseu   = "mg"
)

intervals <- data.frame(
  start    = c(96,   0),
  end      = c(168,  504),
  cmax     = c(TRUE, TRUE),
  cmin     = c(TRUE, FALSE),
  cav      = c(TRUE, FALSE),
  auclast  = c(TRUE, TRUE)
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- suppressWarnings(PKNCA::pk.nca(nca_data))

knitr::kable(
  summary(nca_res),
  caption = "Simulated NCA parameters by cancer type (PKNCA): Cmax / Cmin / Cav over the 96-168 h late-CIVI window and AUClast over the full 21-day cycle."
)
```

## Comparison against the paper's analytic Css

The paper's analytic Css for a typical NSCLC patient with BSA 1.97 m^2 and
CL 42.1 L/h is

`Css = dose_rate / CL = (4.8 * 1.97 / 24) / 42.1 mg/(L) ~ 0.00936 mg/L = 9.36 ng/mL`.

For the median-CRCL melanoma patient with the +24% CL effect:

`Css ~ 9.36 / 1.24 = 7.55 ng/mL`.

For the median-CRCL HRPC patient with the -4.5% CL effect:

`Css ~ 9.36 / 0.955 = 9.80 ng/mL`.

The PKNCA `Cav` (typical 96-168 h window) for each cohort should fall
near these values, modulated by the cohort's CRCL and ALT distributions.

```{r pknca-comparison}
analytic_css <- tibble::tibble(
  cancer            = c("NSCLC", "HRPC", "melanoma"),
  analytic_css_ng_mL = c(
    9.36,
    9.36 / (1 - 0.045),
    9.36 / (1 + 0.24)
  )
)

nca_tbl <- as.data.frame(nca_res$result) |>
  filter(PPTESTCD == "cav", start == 96) |>
  group_by(cancer) |>
  summarise(median_cav_ng_mL = median(PPORRES, na.rm = TRUE),
            .groups = "drop")

knitr::kable(
  analytic_css |>
    left_join(nca_tbl, by = "cancer") |>
    mutate(pct_diff = round(100 * (median_cav_ng_mL / analytic_css_ng_mL - 1), 1)),
  caption = "Analytic Css vs simulated median Cav (96-168 h) by cancer type."
)
```

# Assumptions and deviations

- **Single proportional residual error encoded; the paper's second
  outlier-patient residual is not.** Aoyama 2012 fit two proportional
  residual-error terms to handle 16 anomalous concentrations from 11 of
  96 patients: a typical-patient error (`sigma^2 = 0.0934`, SD = 0.306,
  30.6% CV) and a much larger outlier-patient error (`sigma^2 = 31.6`,
  SD ~ 5.62, applied to the 11 patients with at least one concentration
  exceeding the post-hoc IQR-based threshold `Q3 + 3 * IQR = 23.13
  ng/mL`). The outlier classification is data-driven (a post-hoc
  identification of patients with anomalously high observed
  concentrations during the CIVI), not a population-level predictor that
  a simulation of new patients could mimic. The packaged model therefore
  encodes only the typical-patient residual error; users intending to
  reproduce the full bimodal residual structure for a re-fit on the
  original data would need to introduce a per-patient mixture indicator
  derived from the observed concentrations, which is outside the scope
  of a simulation-oriented model library.

- **Cancer-type effect form: proportional `(1 + e * I)` (mathematically
  identical to the paper's power form for a binary indicator).** The
  paper writes the cancer-type effect as `THETA_HRPC^I_HRPC *
  THETA_MM^I_MM` with `THETA_HRPC = 0.955` and `THETA_MM = 1.24` (Table
  3, "ratio" rows). Because the indicators are binary 0/1, this is
  exactly equivalent to the proportional form `(1 + e_hrpc_cl *
  I_HRPC) * (1 + e_mel_cl * I_MEL)` with `e_hrpc_cl = -0.045` and
  `e_mel_cl = +0.24` used in the model file -- both give factor 1.0 at
  `I = 0` and factor `THETA` at `I = 1`. The proportional form is used
  to match the existing Ahamadi 2017 pembrolizumab convention for
  categorical CL effects (`(1 + e_nsclc_cl * TUMTP_NSCLC)`).

- **NSCLC is the implicit reference category for cancer type.** When
  both `TUMTP_HRPC = 0` and `TUMTP_MEL = 0`, the model collapses to the
  NSCLC baseline. NSCLC is therefore NOT carried as a separate
  `TUMTP_NSCLC` column input to this model (although a downstream user
  decomposing a source TUMTP categorical may still want to materialise
  `TUMTP_NSCLC` for symmetry with other oncology models in the package).

- **Reference covariate values 79 mL/min (CRCL) and 20 U/L (ALT) match
  the published medians from Table 2.** The paper Methods state that
  the reference value `Xpop` is "a median or an arbitrary value close
  to the median"; the medians are used here. The reference values were
  cross-checked against the paper's explicit prediction that
  `CLCR = 40 mL/min` produces a 25% CL reduction:
  `(40/79)^0.425 = 0.749`, i.e. 25.1% reduction, matching the published
  statement.

- **CRCL stored under the canonical `CRCL` column despite NOT being
  BSA-normalized.** Aoyama 2012 uses the raw Cockcroft-Gault equation
  (paper Eq. 5) producing creatinine clearance in mL/min, NOT
  BSA-normalized to mL/min/1.73 m^2. The canonical `CRCL` register
  entry accepts either MDRD/CKD-EPI eGFR or BSA-normalized measured
  CrCl; following the precedent of `Delattre_2010_amikacin.R` and
  `NA_NA_lidocaine.R` (which also use raw Cockcroft-Gault values under
  `CRCL`), the model stores the source `CLCR` column under `CRCL`, with
  the raw / non-BSA-normalized status documented in the per-model
  `covariateData[[CRCL]]$units` and `notes` fields. Reference value
  79 mL/min is paper-derived (Table 2 median); do not compare the
  magnitude of `e_crcl_cl = 0.425` directly against the BSA-normalized
  reference values listed in the canonical entry (80, 90, or 100
  mL/min/1.73 m^2).

- **No IIV on V.** Aoyama 2012 Discussion (page 446-447) explicitly
  notes that the limited terminal-phase sampling (only 0.5-24 h after
  stop of CIVI in cycle 1) made it difficult to separate
  inter-individual variability in V from variability in CL; the final
  model carries IIV on CL only (`omega^2 = 0.0385`). Simulations from
  this model therefore have no between-subject variability in V; the
  spread in late-CIVI concentrations is driven entirely by CL
  variability plus residual error.

- **Two new canonical covariate columns registered alongside this
  extraction.** `TUMTP_HRPC` (hormone-refractory prostate cancer; modern
  term castration-resistant prostate cancer / CRPC) and `TUMTP_MEL`
  (melanoma) follow the existing `TUMTP_NSCLC` / `TUMTP_GC` /
  `TUMTP_SCLC` decomposition pattern. The melanoma indicator uses
  `MEL` (not `MM`) to disambiguate from the existing canonical `MM`
  column for active multiple myeloma; the Aoyama 2012 paper's `MM`
  abbreviation denotes malignant (unresectable) melanoma, NOT multiple
  myeloma. See `inst/references/covariate-columns.md` for the full
  entries.

- **CRPC vs HRPC nomenclature.** The Aoyama 2012 trial cohort is
  labelled `HRPC` (hormone-refractory prostate cancer), which is the
  historical term displaced after ~2008 by `CRPC` (castration-resistant
  prostate cancer); the two terms refer to the same clinical entity.
  The canonical column is registered as `TUMTP_HRPC` and accepts both
  wordings via the source-aliases mechanism.

- **Race / ethnicity not modeled.** Aoyama 2012 Discussion (page 447)
  reports that age, gender, and race were tested as covariates and
  found not to be significant. Race composition (Table 2: African
  American 4.2%, Caucasian-with-Hispanic-detail 45.8%, Caucasian
  Hispanic or Latino 8.3%, Caucasian non-Hispanic or Latino 41.7%) is
  carried in the `population` metadata for reference but is not a
  covariate input to the model.

- **Concentration units (ng/mL) require an explicit *1000 scaling in
  `model()`.** With dose in mg and `vc` in L, the ratio `central / vc`
  carries units of mg/L = 1000 ng/mL. The `Cc` output in the model
  multiplies by 1000 to express concentrations in the paper's reported
  ng/mL units. `checkModelConventions()` issues an info-level message
  flagging this dosing-vs-concentration magnitude mismatch; the
  scaling is intentional and documented in the model file.

- **First-cycle simulation only.** This vignette simulates one 21-day
  cycle (one CIVI plus 14 days off). Extending to multi-cycle dosing is
  mechanically straightforward (add additional dose rows on a 504-hour
  schedule via `ii = 504` and `addl = N - 1`) but is outside the scope
  of the simulation pipeline used for validation here.
