---
title: "hUGT1A1-modRNA for Crigler-Najjar syndrome type 1 (Apgar 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hUGT1A1-modRNA for Crigler-Najjar syndrome type 1 (Apgar 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Apgar and colleagues built a quantitative systems pharmacology (QSP) model of
ALXN1540, a modified messenger RNA encoding human UGT1A1 delivered as a lipid
nanoparticle (LNP), as a treatment for Crigler-Najjar syndrome type 1 (CN1).
The model was calibrated to Gunn rats, the UGT1A1-deficient animal model of
CN1, and then translated to CN1 patients to support first-in-human dose
selection.

The paper presents **two** KroneckerBio model files, and this package therefore
ships two model objects:

```{r models}
rat <- rxode2::rxode(readModelDb("Apgar_2018_ALXN1540_rat_qsp"))
human <- rxode2::rxode(readModelDb("Apgar_2018_ALXN1540_human_qsp"))
```

* Citation: `r rat$reference`
* Article: <https://doi.org/10.1002/psp4.12301>
* Supplementary material (reaction networks, Table S1):
  <https://www.ebi.ac.uk/europepmc/webservices/rest/PMC6391595/supplementaryFiles>

The two files are not a base/final pair: they differ in structure as well as in
parameter values, so each is extracted as published (see *Structural
differences* below).

```{r descriptions, echo = FALSE, results = "asis"}
cat("**Gunn rat model.** ", rat$description, "\n\n", sep = "")
cat("**CN1 human model.** ", human$description, "\n", sep = "")
```

# Population

The Gunn rat model was calibrated against three studies in
Gunn-UGT1a1j/BluHsdRrrc rats: a single-dose 0.3 mg/kg i.v. PK/PD study in adult
(N = 4-5) and juvenile (N = 5) animals; a repeat-dose PD study in adult animals
at 0.1, 0.2 and 0.5 mg/kg Q2W and 0.5 mg/kg Q4W for five doses, with a
0.5 mg/kg Q2W luciferase-mRNA vehicle control (N = 4-5 per time point); and an
escalating single-dose PD study at 0.1, 0.2 and 0.5 mg/kg used as a naive
prediction check (N = 4-6 per time point). Male and female, adolescent and adult
animals were enrolled; because "no clear gender or age differences were noted in
the PK data", all groups were pooled for calibration. Body weights recorded in
the deposited single-dose dataset span 155-262 g (adults) and 70-116 g
(juveniles).

The human model is a **forward projection, not a fit** -- no CN1 patients were
dosed. Biophysical and cellular parameters are carried over from the rat
estimates, and only bilirubin production and clearance are humanized.

```{r population}
str(rat$population, give.attr = FALSE)
str(human$population, give.attr = FALSE)
```

# Source trace

Every `ini()` entry carries an in-file comment naming its source location. The
table below collects them.

| Parameter | Gunn rat | Human | Source |
|---|---|---|---|
| `kw` first-order LNP elimination | 2.41E-05 1/s | 2.41E-05 1/s | Table 1 (Estimated) |
| `k12` LNP central to peripheral | 4.79E-05 1/s | 4.79E-05 1/s | Table 1 (Estimated) |
| `k21` LNP peripheral to central | 2.65E-07 1/s | 2.65E-07 1/s | Table 1 (Estimated) |
| `ka` hepatocyte attachment | 1.17E-05 1/s | 1.17E-05 1/s | Table 1 (Estimated) |
| `ke` endocytosis | 7.70E-05 1/s | 7.70E-05 1/s | Table 1 (Estimated) |
| `de` endosomal degradation | 9.32E-05 1/s | 9.32E-05 1/s | Table 1 (Estimated) |
| `kl` endosomal escape | 1.93E-05 1/s | 1.93E-05 1/s | Table 1 (Estimated) |
| `dmRNA` mRNA degradation | 1.07E-05 1/s | 1.07E-05 1/s | Table 1 (Estimated) |
| `kt` translation | 17.73 1/s | 17.73 1/s | Table 1 (Estimated) |
| `dUGTc` UGT1A1 degradation | 6.76E-06 1/s | 6.76E-06 1/s | Table 1 (Estimated) |
| `kclearBil` | 3.5E-06 1/s | 1.23E-6 1/s | Table 1 (Fixed) |
| `kclearMGT` | 3.5E-05 1/s | 1.23E-5 1/s | Table 1 (Fixed) |
| `kclearDGT` | 3.5E-05 1/s | 1.23E-5 1/s | Table 1 (Fixed) |
| `kon` | 0.001 1/nM/s | 0.001 1/nM/s | Table 1 (Fixed) |
| `kcat` | 0.0011 1/s | 0.0011 1/s | Table 1 (Fixed) |
| `koff` | 0.2589 1/s | 0.2589 1/s | Table 1 (Fixed) |
| `kprod` bilirubin production | Variable | Variable | Table 1; derived from the baseline (below) |
| `ktbg` background translation | not reported | not reported | Supplementary model files, reaction (8); fixed to 0 |
| `ksynhigh`, `kelSbil` | not reported | not reported | Supplementary model files, reactions (15)-(16); fixed to 0 |
| Plasma volume `Vc` | 0.0078 L | 3.0 L | Supplementary model files, `% Compartments` |
| Baseline total bilirubin | 455 nmol | 1.026E+06 nmol (20 mg/dL) | Figure 2A / Figure 5, digitized |
| `dose_scale` nmol per mg/kg | 0.31 | 119.2 | Figure 2A digitized; see *Errata* |

All published rate constants are in 1/s. The model files multiply each by 3600
so the models run on a time base of hours, and keep the published 1/s number
visible inside the `log()` call.

## Reaction network

The reaction network comes from the deposited KroneckerBio files, not from the
main text -- the paper states only that "all reactions are modeled with zero,
first, and second order mass action reactions" and refers the reader to the
supplement.

| # | Reaction | Rate |
|---|---|---|
| 1 | `lnp_central` -> 0; `lnp_central` <-> `lnp_peripheral`; `lnp_peripheral` -> 0 | `kw`, `k12`/`k21`, `kw` |
| 2 | `lnp_central` -> `lnp_attached` | `ka` |
| 3 | `lnp_attached` -> `lnp_endosome` | `ke` |
| 4 | `lnp_endosome` -> 0 | `de` |
| 5 | `lnp_endosome` -> `mrna_cyto` | `kl` |
| 7 | `mrna_cyto` -> 0 | `dmRNA` |
| 8 | 0 -> `ugt_cyto` | `ktbg` (fixed to 0) |
| 9 | `mrna_cyto` -> `mrna_cyto` + `ugt_cyto` | `kt` |
| 10 | `ugt_cyto` -> 0; `bil_ugt` -> `bil` | `dUGTc` |
| 11 | `ugt_cyto` + `bil` <-> `bil_ugt` | `kon`/`koff` |
| 12 | `bil_ugt` -> `ugt_cyto` + `mgt` | `kcat` |
| 13 | rat: `ugt_cyto` + `mgt` -> `ugt_cyto` + `dgt`; human: `ugt_cyto` + `mgt` <-> `ugt_mgt` -> `ugt_cyto` + `dgt` | `kcat`; `kon`/`koff` then `kcat` |
| 14 | 0 -> `bil` | `kprod` |
| 15, 16 | surge production and its decay via `bil_surge` | `ksynhigh`, `kelSbil` (both fixed to 0) |
| 17-19 | `bil` -> 0; `mgt` -> 0; `dgt` -> 0 | `kclearBil`, `kclearMGT`, `kclearDGT` |

## Structural differences between the two published files

Reaction 13 is the one place where the deposited rat and human files genuinely
disagree, and both are reproduced as published:

* The **Gunn rat** file collapses the monoglucuronide step into one second-order
  reaction, `UGTc + MGT -> UGTc + DGT`, reusing `kcat` as its rate constant.
  That makes `kcat` a second-order constant in this one reaction while Table 1
  reports it in 1/s -- a dimensional inconsistency in the source that the
  extraction preserves rather than silently repairs.
* The **CN1 human** file resolves the same step into an explicit `UGTc:MGT`
  enzyme-substrate complex with the same `kon`/`koff`/`kcat` constants, which
  is dimensionally consistent.

The human file additionally carries an optional accelerated-clearance arm
(reactions 20-21, labelled "ADA" with the in-file remark that it "could be
something else"). Its driving input `kada` is set to zero in the deposited file,
neither `kada` nor `kada_deg` appears in Table 1 or anywhere in the paper, and
no published result uses it. It is therefore not encoded; see *Errata*.

## Dimensional analysis

States are amounts in nmol and time is in hours. Second-order reactions are
mass-action in **concentration**, so those rate expressions divide by the plasma
volume:

| Term | Units |
|---|---|
| `kw * lnp_central` | (1/h) x nmol = nmol/h |
| `kon * ugt_cyto * bil / vc` | (1/nM/h) x nmol x nmol / L = (1/nM/h) x nmol x nM = nmol/h |
| `kcat * bil_ugt` | (1/h) x nmol = nmol/h |
| `kcat * ugt_cyto * mgt / vc` (rat only) | (1/h) x nmol x nM -- **not** nmol/h; see the inconsistency noted above |
| `ktbg * vc` | (nmol/L/h) x L = nmol/h |
| `kprod = kclearbil * bl_bilirubin` | (1/h) x nmol = nmol/h |
| `tbili_nmol * mw_bil * 1e-7 / vc` | nmol x (g/mol) x 1e-7 / L = mg/dL |

The mg/dL conversion is `1 nmol = mw_bil * 1e-6 mg`, divided by `vc` litres to
get mg/L and by a further 10 to get mg/dL, i.e. a single factor of
`mw_bil * 1e-7 / vc`.

# Bilirubin production is derived, not tabulated

Table 1 lists `kprod` as "Variable ... Set to match the steady-state bilirubin
levels in the modeled individual or group". Both model files therefore take the
**pre-dose steady-state total bilirubin** as the input parameter
(`bl_bilirubin`, in nmol) and derive `kprod` inside `model()`.

Before dosing there is no UGT1A1 (`ktbg` is fixed to 0, because both CN1 and the
Gunn rat are defined by the absence of UGT1A1 activity), so the only bilirubin
species present is the unconjugated form and

$$\frac{d\,\mathrm{bil}}{dt} = k_{prod} - k_{clearBil}\cdot \mathrm{bil} = 0
\;\;\Longrightarrow\;\; k_{prod} = k_{clearBil}\cdot \mathrm{bl\_bilirubin}.$$

The two published elimination constants are a direct check on the reported
half-lives:

```{r halflives}
hl <- function(k_per_s) log(2) / (k_per_s * 3600)
tibble::tibble(
  Species = c("Gunn rat", "CN1 human"),
  `kclearBil (1/s)` = c(3.5e-6, 1.23e-6),
  `Half-life (h), model` = c(hl(3.5e-6), hl(1.23e-6)),
  `Half-life (h), paper` = c(55, 156)
)
stopifnot(
  abs(hl(3.5e-6) - 55) < 1,
  abs(hl(1.23e-6) - 156) < 1
)
```

# Validation

This is a deterministic mechanistic model with no between-subject variability
and no reported residual error, so the validation follows the endogenous /
mechanistic pattern rather than an NCA comparison. The paper does report that
noncompartmental analysis was performed on the single-dose study, but explicitly
marks those exposure results "data not shown", so there is no published NCA
table to compare against.

```{r helpers}
solve_scenario <- function(mod, dose_mgkg, baseline_nmol,
                           ii_days = 0, n_doses = 1L, end_day = 84,
                           by_h = 6) {
  ev <- rxode2::et(
    amt = dose_mgkg, cmt = "lnp_central",
    ii = ii_days * 24, addl = n_doses - 1L
  )
  ev <- rxode2::et(ev, seq(0, end_day * 24, by = by_h), cmt = "lnp_central")
  out <- suppressWarnings(rxode2::rxSolve(
    mod, ev,
    params = c(bl_bilirubin = baseline_nmol),
    returnType = "data.frame"
  ))
  out$day <- out$time / 24
  out
}

at_day <- function(df, d) {
  v <- df$tbili_nmol[df$day == d]
  if (length(v) != 1L) stop("no unique row at day ", d)
  v
}
```

## 1. Steady-state check

With no dose, both models must hold their reported baseline indefinitely. This
is the check that catches a sign error, a missing elimination term, or a
mistyped baseline.

```{r steady-state}
ss_rat <- solve_scenario(rat, 0, 455, end_day = 180)
ss_hum <- solve_scenario(human, 0, 1.0262e6, end_day = 365)

ss_drift <- c(
  rat = max(abs(ss_rat$tbili_nmol / 455 - 1)),
  human = max(abs(ss_hum$tbili_nmol / 1.0262e6 - 1))
)
ss_drift

# Deterministic solve with a state initialised exactly at its analytic steady
# state: the only error is solver tolerance, so this bound is tight on purpose.
stopifnot(all(ss_drift < 1e-6))

# The human baseline must be 20 mg/dL, the value every panel of Figure 5 starts
# from.
stopifnot(abs(ss_hum$tbili_mgdl[1] - 20) < 0.01)
```

## 2. Perturbation recovery

Displacing total bilirubin away from baseline must return it to baseline, at the
rate set by `kclearBil`.

```{r perturbation, fig.cap = "Perturbation recovery of the Gunn rat model: the unconjugated-bilirubin pool returns to its 455 nmol baseline from both directions."}
pert <- bind_rows(lapply(c(0.5, 1, 2), function(mult) {
  ev <- rxode2::et(seq(0, 30 * 24, by = 6), cmt = "lnp_central")
  s <- rxode2::rxSolve(
    rat, ev,
    params = c(bl_bilirubin = 455),
    inits = c(bil = 455 * mult),
    returnType = "data.frame"
  )
  s$day <- s$time / 24
  s$start <- paste0(mult, "x baseline")
  s
}))

ggplot(pert, aes(day, tbili_nmol, colour = start)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 455, linetype = "dashed") +
  labs(x = "Day", y = "Total bilirubin (nmol)", colour = "Initial state") +
  theme_bw()

final <- pert |>
  filter(day == 30) |>
  pull(tbili_nmol)
# 30 days is ~13 half-lives of unconjugated bilirubin in the Gunn rat.
stopifnot(all(abs(final / 455 - 1) < 0.01))
```

## 3. Mass balance on the bilirubin pool

Summing the four bilirubin-carrying states gives a pool whose only source is
`kprod` and whose only sinks are the three elimination reactions:

$$\frac{d}{dt}\left(\mathrm{bil} + \mathrm{bil\_ugt} + \mathrm{mgt} + \mathrm{dgt}\right)
= k_{prod} - k_{clearBil}\,\mathrm{bil} - k_{clearMGT}\,\mathrm{mgt} - k_{clearDGT}\,\mathrm{dgt}.$$

Every conjugation term cancels, so the integrated balance must close exactly.

```{r mass-balance}
mb <- solve_scenario(rat, 0.3, 455, end_day = 30, by_h = 0.5)

k <- c(bil = 3.5e-6 * 3600, mgt = 3.5e-5 * 3600, dgt = 3.5e-5 * 3600)
kprod <- k[["bil"]] * 455

trap <- function(x, y) sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)

pool <- mb$bil + mb$bil_ugt + mb$mgt + mb$dgt
produced <- kprod * max(mb$time)
eliminated <- trap(mb$time, k[["bil"]] * mb$bil) +
  trap(mb$time, k[["mgt"]] * mb$mgt) +
  trap(mb$time, k[["dgt"]] * mb$dgt)
observed <- tail(pool, 1) - head(pool, 1)

balance <- c(
  produced = produced,
  eliminated = eliminated,
  `net (produced - eliminated)` = produced - eliminated,
  `observed change in pool` = observed,
  `relative closure error` = abs((produced - eliminated - observed) / produced)
)
balance

stopifnot(balance[["relative closure error"]] < 1e-3)

# Mutation control: the gate must go RED if a sink is dropped from the
# accounting. Without this, a balance that closes proves nothing.
eliminated_broken <- trap(mb$time, k[["bil"]] * mb$bil) +
  trap(mb$time, k[["mgt"]] * mb$mgt)
broken_err <- abs((produced - eliminated_broken - observed) / produced)
broken_err
stopifnot(broken_err > 0.1)
```

## 4. The nadir is set by production and diglucuronide clearance

Figure 6 of the paper explains the bilirubin response in three phases and states
that "as the majority of bilirubin is in diglucuronide form the nadir level is
determined by the balance of bilirubin production and the clearance rate of the
diglucuronide". That is a closed-form prediction the model must satisfy:

$$\mathrm{nadir} \approx \frac{k_{prod}}{k_{clearDGT}}
= \frac{k_{clearBil}}{k_{clearDGT}}\cdot \mathrm{bl\_bilirubin}
= \frac{\mathrm{bl\_bilirubin}}{10},$$

because Table 1 fixes `kclearDGT` at ten times `kclearBil` in both species.

```{r nadir}
nadir_check <- bind_rows(lapply(c(455, 600, 1000), function(bl) {
  s <- solve_scenario(rat, 0.5, bl, end_day = 14, by_h = 1)
  tibble::tibble(
    baseline_nmol = bl,
    nadir_nmol = min(s$tbili_nmol),
    predicted_nmol = bl / 10
  )
}))
nadir_check |>
  mutate(pct_diff = 100 * (nadir_nmol - predicted_nmol) / predicted_nmol)

stopifnot(
  all(abs(nadir_check$nadir_nmol / nadir_check$predicted_nmol - 1) < 0.05)
)
```

## 5. Replicating Figure 2 -- single 0.3 mg/kg dose in Gunn rats

The paper's Figure 2 caption states that the adult and juvenile fits differ in
one quantity only: the initial total bilirubin level. Both panels are therefore
the same simulation at two baselines.

```{r fig2, fig.cap = "Replicates Figure 2 of Apgar 2018: plasma mRNA and total bilirubin after a single 0.3 mg/kg i.v. dose in adult and juvenile Gunn rats."}
fig2 <- bind_rows(
  solve_scenario(rat, 0.3, 455, end_day = 3, by_h = 0.25) |> mutate(group = "Adult (455 nmol baseline)"),
  solve_scenario(rat, 0.3, 568, end_day = 3, by_h = 0.25) |> mutate(group = "Juvenile (568 nmol baseline)")
)

fig2 |>
  select(day, group, `Plasma mRNA (nmol)` = plasma_mrna, `Total bilirubin (nmol)` = tbili_nmol) |>
  pivot_longer(-c(day, group)) |>
  ggplot(aes(day, value, colour = group)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~name, scales = "free_y") +
  labs(x = "Time (d)", y = NULL, colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")

# Digitized anchors from Figure 2A (adult panel).
fig2_adult <- filter(fig2, group == "Adult (455 nmol baseline)")
fig2_anchors <- tibble::tibble(
  quantity = c("plasma mRNA at t = 0 (nmol)", "total bilirubin at t = 0 (nmol)",
               "total bilirubin at day 3 (nmol)"),
  model = c(
    fig2_adult$plasma_mrna[fig2_adult$day == 0],
    at_day(fig2_adult, 0),
    at_day(fig2_adult, 3)
  ),
  figure = c(0.093, 455, 50)
)
fig2_anchors |> mutate(pct_diff = 100 * (model - figure) / figure)
```

The initial log-linear slope of the plasma-mRNA curve is an independent check on
three separately-reported rate constants at once: everything leaving the central
LNP compartment is `kw + k12 + ka`.

```{r fig2-slope}
early <- filter(fig2_adult, day > 0, day <= 0.5)
slope_per_day <- coef(lm(log(plasma_mrna) ~ day, data = early))[["day"]]
expected <- -(2.41e-05 + 4.79e-05 + 1.17e-05) * 86400

c(model = slope_per_day, `kw + k12 + ka` = expected)
stopifnot(abs(slope_per_day / expected - 1) < 0.02)
```

## 6. Replicating Figure 3 -- repeat dosing in Gunn rats

```{r fig3, fig.cap = "Replicates Figure 3 of Apgar 2018: total bilirubin under repeat dosing in adult Gunn rats. The luciferase control receives no active drug."}
fig3 <- bind_rows(
  solve_scenario(rat, 0, 1000, end_day = 84, by_h = 6) |> mutate(regimen = "Luciferase control"),
  solve_scenario(rat, 0.1, 1000, ii_days = 14, n_doses = 5L) |> mutate(regimen = "0.1 mg/kg Q2W"),
  solve_scenario(rat, 0.2, 1000, ii_days = 14, n_doses = 5L) |> mutate(regimen = "0.2 mg/kg Q2W"),
  solve_scenario(rat, 0.5, 1000, ii_days = 14, n_doses = 5L) |> mutate(regimen = "0.5 mg/kg Q2W"),
  solve_scenario(rat, 0.5, 1000, ii_days = 28, n_doses = 3L) |> mutate(regimen = "0.5 mg/kg Q4W")
) |>
  mutate(regimen = factor(regimen, levels = c(
    "Luciferase control", "0.1 mg/kg Q2W", "0.2 mg/kg Q2W",
    "0.5 mg/kg Q2W", "0.5 mg/kg Q4W"
  )))

ggplot(fig3, aes(day, tbili_nmol)) +
  geom_line(linewidth = 0.8, colour = "steelblue") +
  facet_wrap(~regimen) +
  labs(x = "Time (d)", y = "Total bilirubin (nmol)") +
  theme_bw()
```

Two qualitative features the paper emphasises are reproduced: the Q4W regimen
shows a deep trough followed by a substantial rebound before the next dose,
whereas Q2W dosing holds the nadir continuously; and after the last dose all
active arms return towards the untreated baseline.

```{r fig3-checks}
q4w <- filter(fig3, regimen == "0.5 mg/kg Q4W")
q2w <- filter(fig3, regimen == "0.5 mg/kg Q2W")

# Q4W rebounds between doses; Q2W does not.
rebound_q4w <- at_day(q4w, 28) / min(q4w$tbili_nmol)
rebound_q2w <- at_day(q2w, 14) / min(q2w$tbili_nmol)
c(`Q4W rebound ratio at day 28` = rebound_q4w,
  `Q2W rebound ratio at day 14` = rebound_q2w)
stopifnot(rebound_q4w > 5, rebound_q2w < 1.5)

# Every active arm returns towards baseline 28 days after the last dose.
stopifnot(at_day(q2w, 84) / 1000 > 0.9)
```

## 7. Replicating Figure 4 -- escalating single dose in Gunn rats

This study was **not** used in calibration; the paper presents it as a naive
prediction check.

```{r fig4, fig.cap = "Replicates Figure 4 of Apgar 2018: total bilirubin after a single escalating i.v. dose in Gunn rats (naive prediction, not used in calibration)."}
fig4 <- bind_rows(lapply(c(0.1, 0.2, 0.5), function(d) {
  solve_scenario(rat, d, 600, end_day = 21, by_h = 3) |>
    mutate(dose = paste0(d, " mg/kg single bolus"))
}))

ggplot(fig4, aes(day, tbili_nmol, colour = dose)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time (d)", y = "Total bilirubin (nmol)", colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

The paper's reading of this study is that "for all three dose levels, a similar
initial reduction of bilirubin is predicted, but dose levels differ in the
predicted duration in which bilirubin levels are held at the nadir level".

```{r fig4-checks}
fig4_summary <- fig4 |>
  group_by(dose) |>
  summarise(
    nadir = min(tbili_nmol),
    # Days spent within 25% of the nadir.
    days_at_nadir = sum(tbili_nmol <= 1.25 * min(tbili_nmol)) * 3 / 24,
    .groups = "drop"
  )
fig4_summary

# Same nadir across doses ...
stopifnot(max(fig4_summary$nadir) / min(fig4_summary$nadir) < 1.05)
# ... but a strictly longer hold at the higher dose.
stopifnot(
  fig4_summary$days_at_nadir[fig4_summary$dose == "0.5 mg/kg single bolus"] >
    fig4_summary$days_at_nadir[fig4_summary$dose == "0.1 mg/kg single bolus"]
)
```

## 8. Replicating Figure 5 -- first-in-human projections

```{r fig5, fig.cap = "Replicates Figure 5a-e of Apgar 2018: projected total bilirubin in CN1 patients for a single dose and for Q4W, Q3W, Q2W and Q1W regimens."}
bl_hum <- 1.0262e6
doses5 <- c(0.005, 0.02, 0.05, 0.2, 0.5)
regimens <- tibble::tribble(
  ~label, ~ii, ~n, ~end,
  "Single Dose", 0, 1L, 12,
  "Q4W", 28, 3L, 84,
  "Q3W", 21, 4L, 84,
  "Q2W", 14, 6L, 84,
  "Q1W", 7, 12L, 84
)

fig5 <- bind_rows(lapply(seq_len(nrow(regimens)), function(i) {
  r <- regimens[i, ]
  bind_rows(lapply(doses5, function(d) {
    solve_scenario(human, d, bl_hum, ii_days = r$ii, n_doses = r$n, end_day = r$end) |>
      mutate(regimen = r$label, dose = paste0(d, " mg/kg"))
  }))
})) |>
  mutate(
    regimen = factor(regimen, levels = regimens$label),
    dose = factor(dose, levels = paste0(doses5, " mg/kg"))
  )

ggplot(fig5, aes(day, tbili_mgdl, colour = dose)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 15, linetype = "dashed") +
  facet_wrap(~regimen, scales = "free_x") +
  labs(x = "Time (d)", y = "Total bilirubin (mg/dL)", colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

Figure 5f plots the trough bilirubin at the end of the dose interval against
dose. The model reproduces its shape, including the strongly supraproportional
gain from shortening the interval.

```{r fig5f, fig.cap = "Replicates Figure 5f of Apgar 2018: projected end-of-interval total bilirubin versus dose for each regimen. The dashed line is the 5 mg/dL reduction benchmark."}
dose_grid <- 10^seq(-5, 0, length.out = 21)

fig5f <- bind_rows(lapply(seq_len(nrow(regimens)), function(i) {
  r <- regimens[i, ]
  read_day <- if (r$label == "Single Dose") 7 else r$end
  bind_rows(lapply(dose_grid, function(d) {
    s <- solve_scenario(human, d, bl_hum, ii_days = r$ii, n_doses = r$n,
                        end_day = r$end, by_h = 12)
    tibble::tibble(
      regimen = r$label, dose = d,
      tbili_mgdl = s$tbili_mgdl[s$day == read_day]
    )
  }))
})) |>
  mutate(regimen = factor(regimen, levels = regimens$label))

ggplot(fig5f, aes(dose, tbili_mgdl, colour = regimen)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 15, linetype = "dashed") +
  scale_x_log10() +
  labs(x = "Dose (mg/kg)", y = "Total bilirubin at end of interval (mg/dL)", colour = NULL) +
  theme_bw()
```

```{r fig5f-doses}
# Dose achieving the paper's 5 mg/dL reduction benchmark (trough = 15 mg/dL),
# by log-linear interpolation on the curves above.
dose_for_15 <- function(reg) {
  d <- filter(fig5f, regimen == reg) |> arrange(dose)
  if (min(d$tbili_mgdl) > 15 || max(d$tbili_mgdl) < 15) return(NA_real_)
  approx(x = d$tbili_mgdl, y = log10(d$dose), xout = 15)$y |> (\(z) 10^z)()
}
bench <- tibble::tibble(
  regimen = regimens$label,
  `model dose for 5 mg/dL reduction (mg/kg)` = vapply(regimens$label, dose_for_15, numeric(1))
)
bench

# The paper states an "~25-fold reduction in the dose required" moving from Q4W
# to Q3W. That RATIO is a scale-free check and is the one the model must meet.
ratio_q4_q3 <- bench[[2]][bench$regimen == "Q4W"] / bench[[2]][bench$regimen == "Q3W"]
ratio_q4_q3
stopifnot(ratio_q4_q3 > 10, ratio_q4_q3 < 100)

# Shortening the interval always lowers the required dose.
ordered_doses <- bench[[2]][match(c("Q4W", "Q3W", "Q2W", "Q1W"), bench$regimen)]
stopifnot(all(diff(ordered_doses) < 0))
```

# Published claims

```{r claims}
claim <- function(text, model_value, threshold_text, pass, deviation = FALSE) {
  tibble::tibble(
    Claim = text, `Model value` = model_value,
    Expected = threshold_text, Pass = pass, Deviation = deviation
  )
}

q4w_h <- solve_scenario(human, 0.5, bl_hum, ii_days = 28, n_doses = 3L, end_day = 84)
sd_h <- solve_scenario(human, 0.005, bl_hum, end_day = 12, by_h = 6)
q1w_h <- solve_scenario(human, 0.005, bl_hum, ii_days = 7, n_doses = 12L, end_day = 84)

claims <- bind_rows(
  claim(
    "Gunn rat unconjugated bilirubin half-life is 55 h",
    sprintf("%.1f h", log(2) / (3.5e-6 * 3600)), "55 h",
    abs(log(2) / (3.5e-6 * 3600) - 55) < 1
  ),
  claim(
    "CN1 unconjugated bilirubin half-life is 156 h",
    sprintf("%.1f h", log(2) / (1.23e-6 * 3600)), "156 h",
    abs(log(2) / (1.23e-6 * 3600) - 156) < 1
  ),
  claim(
    "LNP plasma half-life is ~8 h",
    sprintf("%.1f h", log(2) / (2.41e-5 * 3600)), "~8 h",
    abs(log(2) / (2.41e-5 * 3600) - 8) < 1
  ),
  claim(
    "Liver mRNA degradation half-life is ~18 h",
    sprintf("%.1f h", log(2) / (1.07e-5 * 3600)), "~18 h",
    abs(log(2) / (1.07e-5 * 3600) - 18) < 1
  ),
  claim(
    "Nadir bilirubin is set by production over diglucuronide clearance",
    sprintf("%.0f nmol", min(fig4$tbili_nmol[fig4$dose == "0.5 mg/kg single bolus"])),
    "600 / 10 = 60 nmol",
    abs(min(fig4$tbili_nmol[fig4$dose == "0.5 mg/kg single bolus"]) / 60 - 1) < 0.05
  ),
  claim(
    "0.5 mg/kg Q4W gives a sustained reduction of at least 5 mg/dL",
    sprintf("%.1f mg/dL reduction at trough", 20 - q4w_h$tbili_mgdl[q4w_h$day == 84]),
    "at least 5 mg/dL",
    (20 - q4w_h$tbili_mgdl[q4w_h$day == 84]) >= 5
  ),
  claim(
    "0.005 mg/kg Q1W gives more than a 15 mg/dL reduction",
    sprintf("%.1f mg/dL reduction", 20 - q1w_h$tbili_mgdl[q1w_h$day == 84]),
    "more than 15 mg/dL",
    (20 - q1w_h$tbili_mgdl[q1w_h$day == 84]) > 15
  ),
  claim(
    "Q4W to Q3W needs an ~25-fold lower dose for the same reduction",
    sprintf("%.0f-fold", ratio_q4_q3), "~25-fold",
    ratio_q4_q3 > 10 && ratio_q4_q3 < 100
  ),
  claim(
    "A single 0.005 mg/kg dose gives a transient 5 mg/dL reduction",
    sprintf("%.1f mg/dL reduction at day 7", 20 - sd_h$tbili_mgdl[sd_h$day == 7]),
    "~5 mg/dL",
    FALSE, deviation = TRUE
  )
)
knitr::kable(claims)

stopifnot(all(claims$Pass[!claims$Deviation]))
```

# Assumptions and deviations

## Errata and source inconsistencies

* **`koff` and the quoted `Km` disagree.** Table 1 fixes `koff` at 0.2589 1/s
  and says it was "Computed from Km = 0.20 uM and Km = (koff + kcat)/kon".
  Back-solving with the tabulated `kon` and `kcat` gives
  `Km = (0.2589 + 0.0011) / 0.001 = 260 nM = 0.26 uM`, not 0.20 uM. The
  **tabulated `koff` is used**, because it is the value the published
  simulations ran on; the quoted `Km` is a derivation note.
* **The Q4W-to-Q3W dose ratio is quoted two ways.** The Results state an
  "~25-fold reduction in the dose required" when switching from Q4W to Q3W, but
  name 0.5 mg/kg Q4W and (in the Discussion) 0.002 mg/kg Q3W, which differ
  250-fold. The vignette gates on the stated ~25-fold **ratio**, which the model
  reproduces.
* **The paper also carries two unit slips** that do not affect the model: "a Q1W
  dose of 0.005 mg/dL" (should read mg/kg), and Table 1's "Assumed 10x
  unmodified elimination rate" typesets the multiplication sign such that some
  text extractions render it as an exponent. The tabulated values
  (3.5E-05 versus 3.5E-06) confirm the factor is 10.

## Parameters not reported in any available source

Three constants appear in the deposited KroneckerBio reaction networks but in no
table, figure or sentence of the paper or its supplement. All three are encoded
as `fixed(0)`, which is the documented convention for an unreported parameter,
and each is overridable:

* **`ktbg`, endogenous background UGT1A1 translation** (reaction 8). Fixed to 0.
  This is a derivation rather than a guess: both CN1 and the Gunn rat are
  defined by the absence of UGT1A1 activity, and any non-zero `ktbg` would
  conjugate bilirubin before dosing, contradicting Table 1's own rule that
  `kprod` reproduces a steady state governed by `kclearBil` alone.
* **`ksynhigh` and `kelSbil`, the time-varying bilirubin-production surge**
  (reactions 15-16). Fixed to 0, which leaves the constant production rate the
  paper itself assumed for the adult and escalating-dose studies ("a constant
  bilirubin production rate was assumed"). As a consequence, **the declining
  luciferase-control trajectory of Figure 3 is not reproduced** -- the model
  holds a flat control baseline. The paper fitted that decline with a
  time-varying `kprod` whose values it does not report.
* **The human "ADA" accelerated-clearance arm** (reactions 20-21) is not encoded
  at all: its driving input is zero in the deposited file, neither of its two
  constants is reported, and no published result uses it. Encoding it would have
  required inventing `kada_deg`.

## Dose-scale conversion (the largest source of uncertainty)

The model states are amounts in nmol of LNP-encapsulated modRNA, but every dose
in the paper is in mg/kg, and the paper reports **neither the modRNA molar mass
nor the body weight** used in the simulations. The conversion therefore had to be
recovered:

* **Gunn rat (well determined).** Figure 2A plots the fitted plasma-mRNA curve on
  a calibrated log axis; at `t = 0` for the 0.3 mg/kg dose it reads
  9.3E-02 nmol, i.e. **0.31 nmol per mg/kg**. This digitization is
  cross-validated two ways: the initial log-linear slope of the same curve
  recovers `kw + k12 + ka` to within 1% (gated above), and the implied molar
  mass (0.075 mg into 0.093 nmol for a 250 g rat) is ~800 kDa, which is the
  right size for a ~2400-nucleotide modRNA.
* **CN1 human (poorly determined).** The deposited KroneckerBio files initialise
  every state as `seed x compartment volume`, so a given mg/kg dose enters as the
  same seed *concentration* in both species and the delivered *amount* scales
  with plasma volume. That convention gives
  `0.31 x 3.0 / 0.0078 = 119.2 nmol per mg/kg`, which is the shipped value. An
  alternative route through the rat-implied molar mass and a 70 kg body weight
  gives 87.4 nmol per mg/kg -- only 1.4-fold different.

**Known deviation.** Under either conversion the human model is roughly an order
of magnitude more potent per mg/kg than the paper's stated absolute dose
anchors. The model needs ~0.055 mg/kg Q4W (rather than the published
0.5 mg/kg) and ~0.0006 mg/kg as a single dose (rather than the published
0.005 mg/kg) to produce a 5 mg/dL reduction, and a single 0.005 mg/kg dose
drives bilirubin to ~2 mg/dL instead of the published ~15 mg/dL. The claims
table records that last item as a **Deviation**, excluded from the gate and left
visible.

Two observations bound the problem rather than dissolve it. First, the
*scale-free* claims survive: the ~25-fold Q4W-to-Q3W ratio, the ordering of the
required doses across regimens, the nadir relation, and the exact pre-dose
steady state are all reproduced. Second, the discrepancy is not removed by
switching to the physically-principled molar-mass conversion, so it is unlikely
to be a simple unit error in this extraction; it looks like a reproducibility gap
between Table 1 plus the deposited human reaction network and the published
Figure 5 simulations. The dose conversion was **not** tuned to make Figure 5
match, because that would have made the Figure 5 replication circular.

Users reproducing the paper's human dosing recommendations should set
`dose_scale` explicitly for their purpose; it is a single overridable parameter.

## Other modelling choices

* **Rate units.** All rate constants are published in 1/s and are multiplied by
  3600 so the models run in hours. The published 1/s value is kept visible
  inside each `log()` call, so every `ini()` line is a direct read of Table 1.
* **`kprod` is parameterised by the baseline.** Table 1 defines `kprod` only by
  its effect ("set to match the steady-state bilirubin levels"), so the model
  files take the pre-dose steady-state total bilirubin as the input and derive
  `kprod` from it. Baselines are digitized: 455 nmol (Figure 2A, adult),
  568 nmol (Figure 2B, juvenile), ~1000 nmol (Figure 3) and ~600 nmol
  (Figure 4) in the rat; 20 mg/dL (all panels of Figures 5 and 6) in the human.
* **Liver mRNA and UGT1A1 are reported as total amounts.** The paper's Figure 2
  middle panels plot liver mRNA "per g liver tissue" and Figure 4 plots UGT1A1
  protein on an explicitly arbitrary scale ("an optimal proportionality constant
  was used to relate model output in absolute units to the arbitrary scale of
  the expression data"). Neither the liver mass nor that proportionality
  constant is reported, so `liver_mrna` and `liver_ugt` are left as total nmol
  and those panels are not gated.
* **No NCA comparison.** The paper performed noncompartmental analysis on the
  single-dose study but marks the results "data not shown", so there is no
  published NCA table to compare against. Validation instead follows the
  mechanistic pattern: steady state, perturbation recovery, mass balance,
  dimensional analysis and figure replication.
* **No variability.** The paper reports no between-animal or between-subject
  random effects and no residual-error estimates, stating only that "a 20%
  parameter variation around the estimated or fixed values captured the data
  variability observed in the rat studies". Neither model carries etas or an
  error model; both are intended for simulation rather than estimation.
* **Covariates.** Sex and age group are recorded in `covariatesDataExcluded` on
  the rat model: both were present in the study design but neither is a model
  covariate, because the groups were pooled after the paper found "no clear
  gender or age differences ... in the PK data".
