---
title: "5-Fluorouracil and leukocyte myelosuppression (Arshad 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5-Fluorouracil and leukocyte myelosuppression (Arshad 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Arshad_2020_fluorouracil"))$reference`
* Description: `r rxode2::rxode(readModelDb("Arshad_2020_fluorouracil"))$description`
* Article: <https://doi.org/10.1007/s00280-019-04028-5> (open access; PMC7125253)

Arshad et al. gave 30 patients with gastrointestinal cancer 5-FU 650 or
1000 mg/m^2/day as a 5-day continuous intravenous infusion. The 14 patients
with oesophageal cancer also received cisplatin 20 mg/m^2/day. Plasma 5-FU,
its first catabolite 5-fluoro-5,6-dihydrouracil (5FUH2) and total white blood
cell (WBC) counts were fitted simultaneously in NONMEM. 5-FU follows a
two-compartment model. A fixed 85% of 5-FU clearance forms 5FUH2, which follows
a one-compartment model. A single linear body-surface-area (BSA) effect scales
both clearances. The 5-FU concentration drives a Friberg myelosuppression chain
through a linear drug effect. The slope of that effect is larger in patients
who also received cisplatin.

No supplement or erratum was found for this article (literature check,
2026-09-25).

# Population

```{r population}
str(rxode2::rxode(readModelDb("Arshad_2020_fluorouracil"))$population)
```

Thirty patients (25 men, 5 women) aged 37-73 years (median 59.5) completed the
first cycle (Table 1). Median body weight was 76 kg (46-111) and median BSA was
1.95 m^2 (1.48-2.33). Median baseline WBC count was 6.90 x 10^9/L. Sixteen
patients with colorectal, rectal or anal cancer received 5-FU with radiotherapy.
Fourteen with oesophageal cancer received 5-FU plus cisplatin. The analysis used
199 5-FU and 251 5FUH2 concentrations and 135 WBC counts from 29 patients.

# Source trace

| Model element | Value | Source |
|---|---|---|
| `lcl` (CL5FU at BSA 1.95 m^2) | 249 L/h | Table 3, bootstrap column (NONMEM 256) |
| `lvc` (VC,5FU) | 5.56 L | Table 3, bootstrap (NONMEM 5.85) |
| `lvp` (VP,5FU) | 28.5 L | Table 3, bootstrap (NONMEM 24.0) |
| `lq` (Q) | 14.8 L/h | Table 3, bootstrap (NONMEM 17.3) |
| `e_bsa_cl` (shared by CL5FU and CL5FUH2) | 0.77 per m^2 | Table 3, footnote a; Abstract "77%/m2"; Results (single shared parameter) |
| `fm` | 0.85 (fixed) | Methods, Pharmacokinetic analysis; Table 3 |
| `lcl_5fuh2` | 121 L/h | Table 3, bootstrap (NONMEM 124) |
| `lvc_5fuh2` | 96.7 L | Table 3, bootstrap (NONMEM 100) |
| `lcirc0` | 6.86 x 10^9/L | Table 3, bootstrap (NONMEM 7.16) |
| `lmtt` | 281 h | Table 3, bootstrap (NONMEM 261) |
| `lslope_mono` | 1.17 L/mg | Table 3, bootstrap (NONMEM 1.31) |
| `lslope_comb` | 2.82 L/mg | Table 3, bootstrap (NONMEM 2.10) |
| `gamma` | 0.17 (fixed) | Results, Pharmacodynamic model; Table 3 |
| IIV CL5FU, VC,5FU, CL5FUH2, VC,5FUH2, CIRC0 | 23.0, 145, 28.9, 59.6, 16.4 %CV | Table 3, bootstrap; `omega^2 = log(CV^2 + 1)` |
| Proportional RUV 5-FU, 5FUH2, WBC | sigma^2 0.32, 0.14, 0.08 | Table 3 ("RUV (sigma2)"); SD = sqrt |
| Two-compartment 5-FU, one-compartment 5FUH2 | -- | Results, Pharmacokinetic model; Fig. 1 |
| 5FUH2 formation `fm * CL5FU` | -- | Fig. 1 (`CL5FU x Fm`) |
| BSA form `CL = TVCL * (1 + 0.77 * (BSA - 1.95))` | -- | Table 3 footnote a; Table 1 median BSA |
| Friberg chain, three transits, `kprol = ktr = kcirc`, `MTT = 4/ktr` | -- | Methods, Pharmacodynamic analysis; Fig. 1 |
| Feedback `(Circ0/Circ)^gamma`, effect `kprol * (1 - slope * Cp)` | -- | Methods, Pharmacodynamic analysis |
| Slope by cisplatin group | -- | Results, Covariate relationships; Table 3 |

# Which column of Table 3?

Table 3 prints two sets of estimates: the NONMEM estimates and the bootstrap
medians. They differ by up to 34% (Slopecomb 2.10 vs 2.82 L/mg). The Abstract
and Discussion quote the bootstrap medians throughout (249 L/h, 5.56 L, 28.5 L,
121 L/h, 96.7 L, 6.86 x 10^9/L, 281 h, 2.82 and 1.17 L/mg, 77%/m^2).

The paper's own simulation settles which set the authors used. Figure 3 (left
panel) shows a typical patient on a 5-day infusion, with WBC nadirs of
4.29 x 10^9/L without cisplatin and 2.26 x 10^9/L with it. The curves start at
6.86 x 10^9/L, which is the bootstrap CIRC0 (the NONMEM CIRC0 is 7.16). The
figure does not state the dose. So for each column, the chunk below finds the
daily dose that gives the monotherapy nadir of 4.29. It then predicts the
combination nadir at that same dose. Both arms share every parameter except the
slope, so only a column whose slope ratio is right can hit both nadirs.

```{r column-discriminator}
mod <- readModelDb("Arshad_2020_fluorouracil")
ui <- rxode2::rxode(mod)

nonmem_ui <- ui |>
  rxode2::ini(
    lcl = log(256), lvc = log(5.85), lvp = log(24.0), lq = log(17.3),
    e_bsa_cl = 0.71, lcl_5fuh2 = log(124), lvc_5fuh2 = log(100),
    lcirc0 = log(7.16), lmtt = log(261),
    lslope_mono = log(1.31), lslope_comb = log(2.10)
  )

typical_wbc <- function(model, events) {
  rxode2::rxSolve(model, events, omega = NA, sigma = NA, returnType = "data.frame")
}

five_day_events <- function(daily_mg, cisplatin, end_day = 60) {
  obs_t <- seq(0, end_day * 24, by = 2)
  dplyr::bind_rows(
    data.frame(
      id = 1L, time = 0, amt = daily_mg * 5, rate = daily_mg / 24, evid = 1L,
      cmt = "central", dvid = NA_integer_
    ),
    data.frame(
      id = 1L, time = obs_t, amt = 0, rate = 0, evid = 0L,
      cmt = NA_character_, dvid = 1L
    )
  ) |>
    dplyr::mutate(BSA = 1.95, CONMED_CISPLATIN = cisplatin) |>
    dplyr::arrange(time, dplyr::desc(evid))
}

nadir_of <- function(model, daily_mg, cisplatin) {
  s <- typical_wbc(model, five_day_events(daily_mg, cisplatin))
  i <- which.min(s$WBC)
  c(nadir = s$WBC[i], tnadir_day = s$time[i] / 24)
}

matched <- lapply(list(bootstrap = ui, nonmem = nonmem_ui), function(m) {
  dose <- uniroot(
    function(d) nadir_of(m, d, 0)[["nadir"]] - 4.29,
    interval = c(500, 5000), tol = 0.5
  )$root
  mono <- nadir_of(m, dose, 0)
  comb <- nadir_of(m, dose, 1)
  data.frame(
    daily_dose = dose, nadir_mono = mono[["nadir"]], nadir_comb = comb[["nadir"]],
    tnadir_mono = mono[["tnadir_day"]], tnadir_comb = comb[["tnadir_day"]]
  )
})
disc <- dplyr::bind_rows(matched, .id = "column") |>
  dplyr::mutate(comb_pct_diff = 100 * (nadir_comb - 2.26) / 2.26)

disc |>
  dplyr::mutate(dplyr::across(-column, \(x) signif(x, 3))) |>
  dplyr::rename(
    "Table 3 column" = column,
    "Daily dose matching mono nadir (mg/day)" = daily_dose,
    "Mono nadir (10^9/L)" = nadir_mono,
    "Comb nadir (10^9/L)" = nadir_comb,
    "Mono Tnadir (day)" = tnadir_mono,
    "Comb Tnadir (day)" = tnadir_comb,
    "Comb nadir vs published 2.26 (%)" = comb_pct_diff
  ) |>
  knitr::kable(caption = "Typical patient (BSA 1.95 m^2), 5-day infusion. Published Figure 3: nadirs 4.29 (mono) and 2.26 (comb), Tnadir day 22-25.")

boot_row <- disc[disc$column == "bootstrap", ]
nonmem_row <- disc[disc$column == "nonmem", ]
stopifnot(
  # Deterministic typical-value solves: tight bounds are appropriate.
  abs(boot_row$comb_pct_diff) < 5,
  abs(nonmem_row$comb_pct_diff) > 15,
  boot_row$tnadir_mono >= 22, boot_row$tnadir_comb <= 25
)
```

With the bootstrap column, the dose that gives the 4.29 monotherapy nadir
(about 1750 mg/day, or about 900 mg/m^2/day at 1.95 m^2) also gives the
published combination nadir to within 1%. The nadirs fall on days 23-24, within
the day 22-25 range stated in the Discussion. The NONMEM column cannot hit both
nadirs at any single dose. The packaged model therefore uses the bootstrap
medians, and each `ini()` comment also gives the NONMEM estimate.

# Replicate Figure 3

```{r figure-3-left, fig.width = 6, fig.height = 4}
fig3l <- dplyr::bind_rows(
  typical_wbc(mod, five_day_events(boot_row$daily_dose, 0, 50)) |>
    dplyr::mutate(arm = "5FU mono"),
  typical_wbc(mod, five_day_events(boot_row$daily_dose, 1, 50)) |>
    dplyr::mutate(arm = "5FU + cisplatin")
)
ggplot(fig3l, aes(time / 24, WBC, linetype = arm)) +
  geom_line() +
  coord_cartesian(ylim = c(0, 8)) +
  labs(
    x = "Time after start of infusion (days)", y = "Total WBC count (10^9/L)",
    linetype = NULL,
    caption = sprintf("Replicates Figure 3 (left) of Arshad 2020; 5-day infusion of %.0f mg/day.", boot_row$daily_dose)
  ) +
  theme_bw()
```

The right panel of Figure 3 compares the 5-FU component of one FOLFIRINOX
course (400 mg/m^2 bolus, then 2400 mg/m^2 over 46 h) with one de Gramont day
(300 mg/m^2 bolus, then 300 mg/m^2 over 24 h). The chunk below simulates both
regimens at BSA 1.95 m^2 with each slope.

```{r figure-3-right, fig.width = 6, fig.height = 4}
regimen_events <- function(bolus, infused, dur, cisplatin, bsa = 1.95) {
  obs_t <- seq(0, 100 * 24, by = 4)
  dplyr::bind_rows(
    data.frame(time = 0, amt = bolus * bsa, rate = 0, evid = 1L, cmt = "central", dvid = NA_integer_),
    data.frame(time = 0, amt = infused * bsa, rate = infused * bsa / dur, evid = 1L, cmt = "central", dvid = NA_integer_),
    data.frame(time = obs_t, amt = 0, rate = 0, evid = 0L, cmt = NA_character_, dvid = 1L)
  ) |>
    dplyr::mutate(id = 1L, BSA = bsa, CONMED_CISPLATIN = cisplatin) |>
    dplyr::arrange(time, dplyr::desc(evid))
}
regimens <- tidyr::expand_grid(
  regimen = c("FOLFIRINOX", "de Gramont"), cisplatin = c(0, 1)
) |>
  dplyr::mutate(
    bolus = ifelse(regimen == "FOLFIRINOX", 400, 300),
    infused = ifelse(regimen == "FOLFIRINOX", 2400, 300),
    dur = ifelse(regimen == "FOLFIRINOX", 46, 24)
  )
fig3r <- dplyr::bind_rows(lapply(seq_len(nrow(regimens)), function(i) {
  r <- regimens[i, ]
  typical_wbc(mod, regimen_events(r$bolus, r$infused, r$dur, r$cisplatin)) |>
    dplyr::mutate(
      regimen = r$regimen,
      slope = ifelse(r$cisplatin == 1, "Slope comb", "Slope mono")
    )
}))
fig3r |>
  dplyr::group_by(regimen, slope) |>
  dplyr::summarise(nadir = min(WBC), tnadir_day = time[which.min(WBC)] / 24, .groups = "drop") |>
  dplyr::mutate(published = ifelse(regimen == "FOLFIRINOX", 1.06, 4.20)) |>
  dplyr::mutate(dplyr::across(c(nadir, tnadir_day), \(x) signif(x, 3))) |>
  dplyr::rename(
    "Regimen" = regimen, "Slope used" = slope, "Simulated nadir (10^9/L)" = nadir,
    "Simulated Tnadir (day)" = tnadir_day, "Published nadir (10^9/L)" = published
  ) |>
  knitr::kable(caption = "Figure 3 (right panel) regimens, typical patient, BSA 1.95 m^2.")

ggplot(fig3r, aes(time / 24, WBC, linetype = regimen)) +
  geom_line() +
  facet_wrap(~slope) +
  labs(
    x = "Time after start of infusion (days)", y = "Total WBC count (10^9/L)",
    linetype = NULL, caption = "Compare Figure 3 (right) of Arshad 2020."
  ) +
  theme_bw()
```

The right panel cannot be reproduced from the printed information. The paper
shows nadirs of 1.06 (FOLFIRINOX) and 4.20 (de Gramont) x 10^9/L, and a
rebound to about 9 x 10^9/L. The model gives much shallower nadirs with either
slope. A FOLFIRINOX course delivers about 60% of the 5-FU AUC of the left-panel
5-day infusion, yet the paper shows a deeper nadir for it than the left panel's
combination nadir. So the right panel must have used settings the paper does not
report, such as a different dose basis, BSA or number of administrations. The
left panel, which the text describes in full, is reproduced. The parameters are
not tuned to the right panel.

# Virtual cohort

The cohort has two arms, 5-FU alone and 5-FU plus cisplatin, with 100 virtual
patients each. BSA is drawn from a normal distribution with mean 1.95 m^2 and
SD 0.2, truncated to the observed 1.48-2.33 m^2 (Table 1). The paper does not
report how many patients received 650 vs 1000 mg/m^2/day, so each virtual
patient gets one of the two doses with equal probability.

```{r cohort}
set.seed(20200309)
rxode2::rxSetSeed(20200309)
n_per_arm <- 100
cohort <- tibble::tibble(
  id = seq_len(2 * n_per_arm),
  CONMED_CISPLATIN = rep(c(0, 1), each = n_per_arm),
  BSA = pmin(pmax(rnorm(2 * n_per_arm, 1.95, 0.2), 1.48), 2.33),
  dose_m2 = sample(c(650, 1000), 2 * n_per_arm, replace = TRUE)
) |>
  dplyr::mutate(
    daily_mg = dose_m2 * BSA,
    treatment = ifelse(CONMED_CISPLATIN == 1, "5FU + cisplatin", "5FU mono")
  )

obs_times <- sort(unique(c(
  seq(0, 24, by = 0.5), seq(24, 120, by = 4), 120 + c(5, 30, 60, 90) / 60,
  seq(124, 60 * 24, by = 12)
)))
ev_cohort <- dplyr::bind_rows(
  cohort |>
    dplyr::mutate(
      time = 0, amt = daily_mg * 5, rate = daily_mg / 24, evid = 1L,
      cmt = "central", dvid = NA_integer_
    ),
  tidyr::expand_grid(cohort, time = obs_times) |>
    dplyr::mutate(amt = 0, rate = 0, evid = 0L, cmt = NA_character_, dvid = 1L)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim <- rxode2::rxSolve(
  mod, ev_cohort,
  sigma = NA, returnType = "data.frame",
  keep = c("treatment", "dose_m2", "BSA")
)
```

# Simulated profiles (compare Figure 2)

Figure 2 of the paper is a visual predictive check of 5-FU, 5FUH2 and WBC. The
bands below are the 5th, 50th and 95th percentiles of the individual
predictions of the virtual cohort (residual error not added).

```{r vpc, fig.width = 7, fig.height = 7}
long <- sim |>
  dplyr::select(id, time, treatment, Cc, Cc_5fuh2, WBC) |>
  tidyr::pivot_longer(c(Cc, Cc_5fuh2, WBC), names_to = "output", values_to = "value") |>
  dplyr::mutate(output = dplyr::recode(
    output,
    Cc = "5-FU (mg/L)", Cc_5fuh2 = "5FUH2 (mg/L)", WBC = "WBC (10^9/L)"
  ))
bands <- long |>
  dplyr::filter(output == "WBC (10^9/L)" | time <= 130) |>
  dplyr::group_by(output, treatment, time) |>
  dplyr::summarise(
    p05 = quantile(value, 0.05), p50 = median(value), p95 = quantile(value, 0.95),
    .groups = "drop"
  )
ggplot(bands, aes(time / 24, p50, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.15, colour = NA) +
  geom_line() +
  facet_wrap(~output, scales = "free", ncol = 1) +
  labs(
    x = "Time after start of infusion (days)", y = NULL, colour = NULL, fill = NULL,
    caption = "Median and 90% interval of individual predictions; compare Figure 2 of Arshad 2020."
  ) +
  theme_bw()
```

# PKNCA validation

Table 3 reports the median AUC over 24 h of infusion: 6.72 mg h/L for 5-FU and
12.2 mg h/L for 5FUH2. Both values pool the 650 and 1000 mg/m^2/day patients.
The footnote says this AUC came from an extra integrating compartment in
NONMEM, but it does not say which 24 h. Both species reach steady state within a
few hours of the start of the infusion, apart from patients with a very large
5FUH2 volume. PKNCA therefore computes the 24-48 h AUC, a steady-state dosing
day, for each virtual patient.

```{r pknca}
conc <- sim |>
  dplyr::filter(time <= 48) |>
  dplyr::select(id, time, treatment, Cc, Cc_5fuh2) |>
  tidyr::pivot_longer(c(Cc, Cc_5fuh2), names_to = "analyte", values_to = "conc") |>
  dplyr::filter(!is.na(conc)) |>
  dplyr::mutate(analyte = ifelse(analyte == "Cc", "5FU", "5FUH2"))
dose_df <- tidyr::expand_grid(
  cohort |> dplyr::select(id, treatment, daily_mg),
  analyte = c("5FU", "5FUH2")
) |>
  dplyr::mutate(time = 0, amt = daily_mg * 5, duration = 120)

conc_obj <- PKNCA::PKNCAconc(conc, conc ~ time | treatment + analyte + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + analyte + id, duration = "duration")
intervals <- data.frame(start = 24, end = 48, auclast = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_tbl <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::group_by(analyte) |>
  dplyr::summarise(
    median = median(PPORRES), p05 = quantile(PPORRES, 0.05), p95 = quantile(PPORRES, 0.95),
    .groups = "drop"
  )
nca_tbl |>
  dplyr::mutate(dplyr::across(-analyte, \(x) signif(x, 3))) |>
  dplyr::rename(
    "Analyte" = analyte, "Median AUC24-48 (mg h/L)" = median,
    "5th percentile" = p05, "95th percentile" = p95
  ) |>
  knitr::kable(caption = "PKNCA AUC over the second infusion day (24-48 h), both arms pooled.")
```

## Comparison against published values

```{r nca-compare}
published_nca <- tibble::tibble(
  analyte = c("5FU", "5FUH2"),
  auclast = c(6.72, 12.2)
)
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published_nca,
  by = "analyte",
  params = "auclast",
  units = c(auclast = "mg*h/L"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Simulated median AUC24-48 vs Table 3 of Arshad 2020. * differs from reference by >20%.")
```

The simulated 5-FU AUC depends on the unreported split between the 650 and
1000 mg/m^2/day patients. The ratio of 5FUH2 to 5-FU AUC does not. With a mass
basis it equals `fm * CL5FU / CL5FUH2` for each patient, and the BSA effect
cancels because both clearances share it. For the typical patient that is
0.85 x 249 / 121 = 1.75, against 12.2 / 6.72 = 1.82 from the Table 3 medians.

```{r auc-ratio}
ratio <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(id, analyte, PPORRES) |>
  tidyr::pivot_wider(names_from = analyte, values_from = PPORRES) |>
  dplyr::mutate(ratio = `5FUH2` / `5FU`)
published_ratio <- 12.2 / 6.72
ratio_summary <- c(
  simulated_median = median(ratio$ratio),
  typical_closed_form = 0.85 * 249 / 121,
  published_median_ratio = published_ratio
)
signif(ratio_summary, 3)
stopifnot(
  # Deterministic: the typical steady-state ratio against the published one.
  abs(ratio_summary[["typical_closed_form"]] / published_ratio - 1) < 0.05,
  # Cohort centre: the median of exp(etalcl - etalcl_5fuh2) is 1, and its
  # sampling SE with 200 patients is about 3%, so 15% is a wide envelope.
  abs(median(ratio$ratio) / ratio_summary[["typical_closed_form"]] - 1) < 0.15
)
```

The typical ratio is 4% below the ratio of the two published medians. The
cohort median lies a few percent lower again, because of the random draws of
the two clearance etas. The two Table 3 AUC medians are taken separately over
patients, so their quotient is only an approximate reference.

# Assumptions and deviations

* **Bootstrap medians, not NONMEM estimates.** Table 3 prints both. The
  packaged values are the bootstrap medians. These are the values the Abstract
  and Discussion quote, and they reproduce the Figure 3 simulation (see "Which
  column of Table 3?"). The NONMEM estimates are recorded in the `ini()`
  comments.
* **IIV scale.** Table 3 reports IIV as %CV without stating the conversion.
  The model uses `omega^2 = log(CV^2 + 1)`. With `omega = CV` instead, the
  145% CV on VC,5FU would give omega^2 = 2.10 rather than 1.13. This affects the
  width of the peak-concentration spread only, not the typical values.
* **Residual error scale.** The Table 3 header says "RUV (sigma2)", so the
  values are variances and the model uses their square roots. The NONMEM RSEs
  (10.2%, 8.06% and 8.70%) are close to the `sqrt(2/N)` expected for a variance
  estimated from 199, 251 and 135 observations. That is consistent with this
  reading.
* **5FUH2 on a mass basis.** The paper says PK parameters used the absolute dose
  and gives no molecular-weight correction. The model feeds `fm * CL5FU * Cc`
  (mg) directly into the 5FUH2 compartment. The molar-mass ratio of 5FUH2 to
  5-FU is 1.015, so a molar reading would change 5FUH2 concentrations by 1.5%.
* **Shared BSA effect.** The paper estimated a single BSA coefficient for both
  clearances. It is encoded as one parameter, `e_bsa_cl`, used in both
  equations. The BSA formula is not reported.
* **Cisplatin indicator.** The paper estimated one slope per group. Here
  `CONMED_CISPLATIN` selects between them. In the study the cisplatin group is
  also the oesophageal-cancer group, so the indicator cannot separate a
  cisplatin effect from a tumour-site effect.
* **Dose split in the virtual cohort.** The number of patients on 650 vs
  1000 mg/m^2/day is not reported, so the virtual cohort assigns them with equal
  probability.
* **Figure 3 right panel** is not reproducible from the stated regimens (see
  above), and the parameters were not adjusted towards it.
* **DPYD, TS and MTHFR genotypes** were tested but not retained in the final
  model (Results, Covariate relationships). They are not included.
