---
title: "Iohexol (Asberg 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Iohexol (Asberg 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Asberg_2020_iohexol"))$reference`
* Description: `r rxode2::rxode(readModelDb("Asberg_2020_iohexol"))$description`
* Article (open access): <https://doi.org/10.1016/j.ekir.2019.11.012>
* Supplement (Figures S1-S3): <https://europepmc.org/article/PMC/PMC7000849#supplementary-material>

Iohexol is a non-ionic X-ray contrast agent that is eliminated almost
entirely by glomerular filtration, so its serum clearance *is* the measured
glomerular filtration rate (GFR). The routine clinical method fits a
log-linear line through two elimination-phase samples (2 h and 5 h, delayed
to 8 h or 24 h when GFR is low) and applies the Brochner-Mortensen correction
for the missed distribution phase. Asberg 2020 replaces that with a
two-compartment population model used as a Bayesian prior, so that GFR can be
read off the individual clearance estimate from four samples within 5 hours
(10 min, 30 min, 2 h, 5 h) regardless of renal function.

# Population

The development cohort was 176 patients aged 1-82 years (Table 1: 2 aged
0-2 years, 38 aged 2-21, 63 aged 21-60 and 73 over 60), mainly Caucasian and
about 76% male, pooled from a prospective Oslo study (2014-2017) and two
previously published cohorts from Tenerife and Rochester. Development-cohort
mean weight by age bin was 11.4, 45.0, 78.9 and 87.9 kg and mean plasma
creatinine 22, 98, 276 and 249 umol/L, i.e. the adult patients were weighted
heavily toward reduced renal function. Across development and validation
cohorts (219 patients) measured GFR spanned 14-149 mL/min, reported as
absolute mL/min rather than per 1.73 m^2.

Each patient received a single IV bolus of 5 mL Omnipaque 300 mg I/mL
(3235 mg iohexol; 2 mL in children under 2 years), with the exact dose found
by weighing the syringe. The development cohort contributed 1131 serum
concentrations (median 7 per patient, range 2-12), measured by HPLC-UV with a
lower limit of quantification of 20 mg/L. A validation cohort of 43 patients
(395 concentrations) was held out for the limited-sampling evaluation.

# Source trace

| Model element | Value | Source location |
|---|---|---|
| Structure: 2-compartment, IV bolus, parameterised as CL, Q, V, Vp | -- | Results, "Model Development" (2-cmt chosen over 1- and 3-cmt) |
| `lcl` | log(1.55 L/h) at 85 kg | Table 2, CLs median |
| `lq` | log(5.99 L/h) at 85 kg | Table 2, Qs median |
| `lvc` | log(10.47 L) at 85 kg | Table 2, Vs median |
| `lvp` | log(8.02 L) at 85 kg | Table 2, Vps median |
| `e_wt_cl_q` | 0.75 (fixed) | Table 2 footnote a; Results final-model equations |
| `e_wt_vc_vp` | 1 (fixed) | Table 2 footnote a; Results final-model equations |
| Reference weight | 85 kg | Table 2 footnote a; Methods "centralized to population median values" |
| `addSd`, `propSd`, `quadSd` | 0.1523073, 0.01747435, -0.000003919581 | Methods, "Model Development" (assay SD polynomial) |
| `gammaSd` | 1.947 | Results, "Model Development" ("The final g value was 1.947") |
| Residual model: error = SD * gamma | -- | Methods, "Model Development" (gamma error model) |
| IIV | not encoded | Table 2 reports mean / median / 95% CI only; Figure S1 (see Assumptions) |

# The typical profile

```{r mod}
mod <- rxode2::rxode(readModelDb("Asberg_2020_iohexol"))
dose_std <- 3235 # mg, 5 mL Omnipaque 300 (Methods)

# The typical 85 kg patient, with Table 2 medians.
p <- list(cl = 1.55, q = 5.99, vc = 10.47, vp = 8.02)
```

The closed-form bi-exponential solution for an IV bolus into the central
compartment is the reference for the ODE: both sides use the same parameters,
so the difference is pure integration error and a tight bound is correct.

```{r closed-form}
biexp <- function(t, dose, cl, q, vc, vp) {
  k <- cl / vc
  k12 <- q / vc
  k21 <- q / vp
  s <- k + k12 + k21
  alpha <- (s + sqrt(s^2 - 4 * k * k21)) / 2
  beta <- (s - sqrt(s^2 - 4 * k * k21)) / 2
  A <- dose / vc * (alpha - k21) / (alpha - beta)
  B <- dose / vc * (k21 - beta) / (alpha - beta)
  A * exp(-alpha * t) + B * exp(-beta * t)
}

tgrid <- c(0.05, 0.167, 0.5, 1, 2, 3, 5, 8, 12, 24)
ev_typ <- data.frame(
  id = 1L,
  time = c(0, tgrid),
  amt = c(dose_std, rep(0, length(tgrid))),
  evid = c(1L, rep(0L, length(tgrid))),
  cmt = "central",
  WT = 85
)
sim_typ <- rxode2::rxSolve(
  mod, ev_typ,
  rtol = 1e-10, atol = 1e-12, returnType = "data.frame"
)
cf <- biexp(sim_typ$time, dose_std, p$cl, p$q, p$vc, p$vp)
rel_err <- max(abs(sim_typ$Cc / cf - 1))
rel_err
# Measured 5e-15 with rxode2 5.1.8 (which still converts to linCmt()); on
# the numeric ODE path (useLinCmt = FALSE) at rtol = 1e-10 it measured
# 1.8e-10. The bound is >100x above the numeric floor so either path passes.
stopifnot(rel_err < 1e-7)
```

# Replicating the supplement's pcVPC (Figure S2)

Figure S2 is a prediction-corrected VPC of the development cohort. Its red
line is the prediction-corrected **observed** median, which after prediction
correction sits on the scale of the median-covariate population prediction --
here the 85 kg population-median patient given the standard 3235 mg dose. The
line was digitised at its bin vertices from the supplement PDF (axis ticks
located from the PDF text layer, red pixels traced at 200 dpi).

Table 2 prints both a weighted **mean** and a weighted **median** for every
parameter, and they differ 1.7-fold for CL and 1.6-fold for Q, so which column
is the typical value is not a formality. Both are overlaid:

```{r pcvpc}
fig_s2 <- data.frame(
  time = c(0.27, 1, 2, 3, 4.87, 6, 7, 24),
  Cc_pub = c(276, 191, 149, 126, 98, 98, 93, 28)
)

p_mean <- list(cl = 2.60, q = 9.44, vc = 10.96, vp = 9.44) # Table 2 'Mean'

ev_s2 <- data.frame(
  id = 1L,
  time = c(0, fig_s2$time),
  amt = c(dose_std, rep(0, nrow(fig_s2))),
  evid = c(1L, rep(0L, nrow(fig_s2))),
  cmt = "central",
  WT = 85
)
sim_s2 <- rxode2::rxSolve(mod, ev_s2, returnType = "data.frame")

cmp_s2 <- fig_s2 |>
  mutate(
    Cc_median_params = sim_s2$Cc,
    Cc_mean_params = biexp(time, dose_std, p_mean$cl, p_mean$q, p_mean$vc, p_mean$vp),
    pct_diff_median = 100 * (Cc_median_params / Cc_pub - 1),
    pct_diff_mean = 100 * (Cc_mean_params / Cc_pub - 1)
  )

cmp_s2 |>
  mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
  rename(
    "Time (h)" = time,
    "Figure S2 median (mg/L)" = Cc_pub,
    "Model, Table 2 medians (mg/L)" = Cc_median_params,
    "Table 2 means (mg/L)" = Cc_mean_params,
    "% diff, medians" = pct_diff_median,
    "% diff, means" = pct_diff_mean
  ) |>
  knitr::kable()
```

```{r pcvpc-plot, fig.cap = "Replicates the observed-median line of Figure S2 of Asberg 2020 (points, digitised). Solid line: this model (Table 2 medians, 85 kg, 3235 mg). Dashed line: the Table 2 means, for contrast."}
tdense <- seq(0.1, 24, by = 0.1)
curves <- bind_rows(
  data.frame(time = tdense, Cc = biexp(tdense, dose_std, p$cl, p$q, p$vc, p$vp),
             params = "Table 2 medians (encoded)"),
  data.frame(time = tdense, Cc = biexp(tdense, dose_std, p_mean$cl, p_mean$q, p_mean$vc, p_mean$vp),
             params = "Table 2 means")
)
ggplot() +
  geom_line(data = curves, aes(time, Cc, linetype = params)) +
  geom_point(data = fig_s2, aes(time, Cc_pub), colour = "red", size = 2) +
  labs(x = "Time after dose (h)", y = "Iohexol (mg/L)", linetype = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r pcvpc-gate}
# The encoded medians track the published median within digitisation and
# binning error (the 24 h bin pools samples drawn over several hours); the
# means undershoot the terminal phase threefold. This is what settles the
# choice of column, and it is deterministic (no simulated cohort).
stopifnot(
  median(abs(cmp_s2$pct_diff_median)) < 10,
  max(abs(cmp_s2$pct_diff_median)) < 25,
  cmp_s2$pct_diff_mean[cmp_s2$time == 24] < -50
)
```

# Body-weight scaling

The only covariate is total body weight, entering all four parameters as
(WT/85) with exponents 0.75 on CL and Q and 1 on V and Vp. A small virtual
set spanning the development cohort's age-bin mean weights shows the
consequence: the initial concentration scales inversely with weight
(volume ~ WT), while the terminal half-life shortens in smaller patients
(CL ~ WT^0.75 against V ~ WT).

```{r weights}
wts <- c(11.4, 45, 78.9, 87.9) # Table 1 development-cohort age-bin means
doses <- c(3235 * 2 / 5, 3235, 3235, 3235) # 2 mL under 2 years (Methods)
tobs <- c(0, 0.083, 0.167, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 5, 6, 8, 12, 16, 24, 36, 48)
ev_wt <- bind_rows(lapply(seq_along(wts), function(i) {
  data.frame(
    id = i,
    time = c(0, tobs),
    amt = c(doses[i], rep(0, length(tobs))),
    evid = c(1L, rep(0L, length(tobs))),
    cmt = "central",
    WT = wts[i]
  )
}))
sim_wt <- rxode2::rxSolve(
  mod, ev_wt,
  rtol = 1e-10, atol = 1e-12,
  returnType = "data.frame", keep = "WT"
) |>
  mutate(group = paste0(WT, " kg"))
```

```{r weights-plot, fig.cap = "Typical-value iohexol profiles at the Table 1 development-cohort age-bin mean weights (the 11.4 kg child receives the 2 mL dose)."}
ggplot(sim_wt, aes(time, Cc, colour = group)) +
  geom_line() +
  geom_hline(yintercept = 20, linetype = "dotted") +
  scale_y_log10() +
  labs(x = "Time after dose (h)", y = "Iohexol (mg/L, log scale)", colour = "Weight",
       caption = "Dotted line: HPLC-UV LLOQ 20 mg/L") +
  theme_bw()
```

# PKNCA validation

Asberg 2020 reports no NCA table, so the NCA check here is whether PKNCA on
the typical profiles recovers the clearance (= GFR) and steady-state volume
implied by Table 2 after weight scaling -- CL = 1.55 (WT/85)^0.75 L/h and
Vss = (10.47 + 8.02) (WT/85) L. Profiles run to 48 h and are left uncensored
so AUC extrapolation is small.

```{r pknca}
conc <- sim_wt |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, group)
dose_df <- ev_wt |>
  filter(evid == 1) |>
  mutate(group = paste0(WT, " kg")) |>
  select(id, time, amt, group)

conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ time | group + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | group + id, route = "intravascular")
intervals <- data.frame(
  start = 0, end = Inf,
  cl.obs = TRUE, vss.obs = TRUE, aucinf.obs = TRUE, half.life = TRUE, c0 = TRUE
)
nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

reference <- data.frame(
  group = paste0(wts, " kg"),
  cl.obs = 1.55 * (wts / 85)^0.75,
  vss.obs = (10.47 + 8.02) * (wts / 85)
)
tbl <- ncaComparisonTable(
  nca, reference,
  by = "group",
  params = c("cl.obs", "vss.obs"),
  units = c(cl.obs = "L/h", vss.obs = "L")
)
knitr::kable(tbl, caption = "Reference: Table 2 medians scaled by (WT/85)^0.75 (CL) and (WT/85) (Vss).")

nca_res <- as.data.frame(nca$result)
cl_nca <- nca_res |> filter(PPTESTCD == "cl.obs") |> arrange(match(group, reference$group))
vss_nca <- nca_res |> filter(PPTESTCD == "vss.obs") |> arrange(match(group, reference$group))
# Linear-up/log-down trapezoids on a 19-point grid; the error is dominated by
# the distribution-phase curvature, measured at < 1% here.
stopifnot(
  max(abs(cl_nca$PPORRES / reference$cl.obs - 1)) < 0.02,
  max(abs(vss_nca$PPORRES / reference$vss.obs - 1)) < 0.05
)
```

# The clinical 2-point method as an independent check

The paper's comparator, GFR_Clin2, fits a log-linear line through two
elimination-phase samples, CL1 = dose / (c1 / b1), and corrects it for the
missed distribution phase with GFR_Clin2 = CL1 / (1 + f CL1),
f = 0.0032 BSA^-1.3 (Methods; CL1 in mL/min). The second sample is taken at
5 h when eGFR > 40, 8 h when < 40 and 24 h when < 30 mL/min per 1.73 m^2. In
the validation cohort GFR_Clin2 agreed with the model-based reference GFR to
a relative predictive error of -0.2 +/- 7.1% (Table 3), with 91% of values
within 10%.

That agreement is informative about the *distribution* parameters, which the
empirical Brochner-Mortensen correction knows nothing about: if Q, V or Vp
were mis-transcribed, the two-point intercept and slope would be biased and
the corrected value would drift away from the true clearance. Applying the
clinical procedure to noise-free typical-patient profiles, sweeping the true
GFR across the paper's 14-149 mL/min range, is therefore a check that uses a
method entirely external to the model. BSA is taken as 2.0 m^2
(Mosteller, 85 kg and 170 cm; Table 1 adult mean heights are 170-179 cm).

```{r clin2}
gfr_true <- c(15, 20, 25, 30, 35, 40, 50, 60, 80, 100, 120, 150) # mL/min
bsa <- sqrt(170 * 85 / 3600)
t2 <- ifelse(gfr_true < 30, 24, ifelse(gfr_true < 40, 8, 5)) # 85 kg: eGFR/1.73 ~ GFR
ev_c2 <- bind_rows(lapply(seq_along(gfr_true), function(i) {
  data.frame(
    id = i, time = c(0, 2, t2[i]), amt = c(dose_std, 0, 0),
    evid = c(1L, 0L, 0L), cmt = "central", WT = 85
  )
}))
# Per-subject clearance override: the model is typical-value, so lcl is set
# per id through the params table to realise each true GFR.
prm <- data.frame(id = seq_along(gfr_true), lcl = log(gfr_true * 60 / 1000))
sim_c2 <- rxode2::rxSolve(mod, ev_c2, params = prm, returnType = "data.frame")

clin2 <- sim_c2 |>
  group_by(id) |>
  arrange(time, .by_group = TRUE) |>
  summarise(
    b1 = log(Cc[1] / Cc[2]) / (time[2] - time[1]), # 1/h
    c1 = Cc[1] * exp(b1 * time[1]), # mg/L
    .groups = "drop"
  ) |>
  mutate(
    gfr_true = gfr_true,
    cl1 = dose_std / (c1 / b1) * 1000 / 60, # mL/min
    f = 0.0032 * bsa^-1.3,
    gfr_clin2 = cl1 / (1 + f * cl1),
    pct_diff = 100 * (gfr_clin2 / gfr_true - 1)
  )
clin2 |>
  transmute(
    "True GFR (mL/min)" = gfr_true,
    "Second sample (h)" = t2,
    "CL1, uncorrected (mL/min)" = signif(cl1, 3),
    "GFR_Clin2 (mL/min)" = signif(gfr_clin2, 3),
    "% diff" = signif(pct_diff, 2)
  ) |>
  knitr::kable()
```

The typical patient sits a few percent above the true GFR (median about +6%,
range about -2% to +10%), i.e. inside one standard deviation of the paper's
-0.2 +/- 7.1% and inside its 10% agreement band. A small positive offset at
the *median* distribution parameters is expected: the Table 2 medians of Q
and Vp are well below their means, and slower distribution is exactly what
makes a two-point line over-read clearance.

The same procedure also reproduces the paper's motivating problem. If the
second sample is always taken at 5 h, the Brochner-Mortensen value drifts
upward as GFR falls, because at low GFR a 2-5 h line is still inside the
distribution phase -- which is why the clinical protocol delays the second
sample to 8 or 24 h, and why Asberg 2020 wanted a method that does not.

```{r clin2-5h}
ev_5h <- ev_c2 |>
  group_by(id) |>
  mutate(time = c(0, 2, 5)) |>
  ungroup()
sim_5h <- rxode2::rxSolve(mod, ev_5h, params = prm, returnType = "data.frame")
clin2_5h <- sim_5h |>
  group_by(id) |>
  arrange(time, .by_group = TRUE) |>
  summarise(
    b1 = log(Cc[1] / Cc[2]) / (time[2] - time[1]),
    c1 = Cc[1] * exp(b1 * time[1]),
    .groups = "drop"
  ) |>
  mutate(
    gfr_true = gfr_true,
    cl1 = dose_std / (c1 / b1) * 1000 / 60,
    gfr_clin2 = cl1 / (1 + 0.0032 * bsa^-1.3 * cl1),
    pct_diff = 100 * (gfr_clin2 / gfr_true - 1)
  )
clin2_5h |>
  transmute(
    "True GFR (mL/min)" = gfr_true,
    "GFR_Clin2, 2 h + 5 h (mL/min)" = signif(gfr_clin2, 3),
    "% diff" = signif(pct_diff, 2)
  ) |>
  knitr::kable()
```

```{r clin2-gate}
# The paper's own Clin2-vs-reference agreement is -0.2 +/- 7.1% with 91% of
# patients within 10% (Table 3). The typical-patient profiles, sampled per the
# clinical protocol, must sit inside that envelope (measured: median +6.4%,
# max |diff| 9.6%). A mis-read distribution parameter moves the uncorrected
# CL1 by tens of percent and would break this. With the second sample fixed at
# 5 h the error must instead grow as GFR falls (measured +29% at 15 mL/min),
# the paper's stated reason for delayed sampling. Deterministic (no cohort).
stopifnot(
  abs(median(clin2$pct_diff)) < 7.1,
  max(abs(clin2$pct_diff)) < 12,
  clin2_5h$pct_diff[gfr_true == 15] > 20,
  all(diff(clin2_5h$pct_diff[gfr_true <= 40]) < 0)
)
```

# Assumptions and deviations

* **Typical values are the Table 2 medians, not the means.** Table 2 prints
  both, weighted by support-point probability, and gives no guidance on which
  is "typical". The medians reproduce the supplement's pcVPC observed median
  within a few percent across 0.3-24 h, while the means undershoot the 24 h
  concentration threefold (see above). The medians are also the natural
  choice for a log-normal-style parameterisation (`exp(lcl)` is a median).
* **No between-subject variability is encoded.** NPAG estimates a discrete
  joint distribution (151 support points), and the paper publishes no SD,
  CV% or variance. The 95% CI column brackets the *mean* (CL 2.22-2.97 around
  2.60, excluding the median 1.55), so it is precision of the mean, not
  between-subject spread. Figure S1 shows marginal support-point densities
  that are flat, bounded and piled up at the search-grid limits (CL
  0.2-10 L/h, V 1-20 L, Vp 1-25 L, Q 0-55 L/h), which no log-normal
  describes. Users who need a stochastic prior can consider two lognormal
  reconstructions, computed below; they agree for Q but disagree about
  twofold for V and CL, which is why neither is encoded:

```{r iiv-reconstruct}
tab2 <- data.frame(
  parameter = c("CL", "Q", "V", "Vp"),
  mean = c(2.60, 9.44, 10.96, 9.44),
  median = c(1.55, 5.99, 10.47, 8.02),
  lo = c(2.22, 8.03, 10.25, 8.66),
  hi = c(2.97, 11.42, 11.68, 10.23)
)
tab2 |>
  mutate(
    omega2_mean_median = 2 * log(mean / median),
    sd_from_ci = (hi - lo) / (2 * 1.96) * sqrt(176),
    omega2_ci = log(1 + (sd_from_ci / mean)^2)
  ) |>
  mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
  rename(
    "Parameter" = parameter,
    "Mean" = mean,
    "Median" = median,
    "95% CI low" = lo,
    "95% CI high" = hi,
    "omega^2 from mean/median" = omega2_mean_median,
    "SD from CI width (n = 176)" = sd_from_ci,
    "omega^2 from CI width" = omega2_ci
  ) |>
  knitr::kable()
```

* **Residual error.** The Pmetrics gamma model multiplies the published assay
  SD polynomial by the fitted gamma (1.947). Pmetrics evaluates that
  polynomial on the *observed* concentration; nlmixr2 evaluates it on the
  prediction (`Cc`). The quadratic coefficient is negative, so the SD peaks
  near 2230 mg/L -- well above the 20-1100 mg/L validated range and above any
  concentration a 3235 mg bolus produces in an adult.
* **The creatinine covariate model is not encoded.** CL = CLs WTc^0.75
  exp(CLCRE CREAT) was tested and rejected because it worsened individual
  predictions; the paper does not print CLCRE.
* **Weight range.** The development cohort included two children under
  2 years; the authors regard the model as validated from 5 to 77 years.
* **Limited-sampling results (Table 3) are not reproduced.** They are Bayesian
  posterior estimates in individual patients against the full non-parametric
  prior, which a typical-value model cannot reproduce.
* **The clinical 2-point check** assumes BSA 2.0 m^2 and uses the eGFR bands
  on the true absolute GFR (for an 85 kg, 2.0 m^2 adult, mL/min and
  mL/min per 1.73 m^2 differ by about 14%, which only moves the 40 and
  30 mL/min thresholds, not the correction).
