---
title: "Benznidazole in T. cruzi-infected mice (Assmus 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benznidazole in T. cruzi-infected mice (Assmus 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Assmus_2025_benznidazole_mouse"))$reference`
* Description: `r rxode2::rxode(readModelDb("Assmus_2025_benznidazole_mouse"))$description`
* Article: <https://doi.org/10.1371/journal.pntd.0012968>
* Supporting information (S1 Code holds the NONMEM control stream of the final PK model):
  <https://journals.plos.org/plosntds/article?id=10.1371/journal.pntd.0012968#sec023>

This is a preclinical, two-layer model. A population PK layer was estimated in
uninfected satellite mice; the resulting typical profiles were then used to
generate exposure metrics that drive a binary logistic exposure-response layer
for sterile parasitological cure in a separate cohort of chronically infected
mice.

# Population

The **PK layer** was estimated from 52 uninfected female BALB/c mice given a
single oral gavage dose of benznidazole at 10 mg/kg (n = 16), 30 mg/kg (n = 18)
or 100 mg/kg (n = 18). Mice weighed 18.1-21.6 g at dosing (median 19.4 g,
SD 0.92 g). Sampling was sparse: 1-3 samples per mouse at nominal times of 0.25,
0.5, 1, 2, 4, 4.5, 6, 8, 10, 12, 14, 16 and 24 h. Of 110 collected samples, two
outliers were excluded and data were censored beyond 8 h (10 mg/kg) and 12 h
(30 mg/kg) because all later concentrations were below the 5 ng/mL LLOQ; 90
quantifiable samples entered the final fit. Estimation used NONMEM 7.4 with
FOCE-I on natural-log-transformed concentrations, and parameter precision came
from sampling importance resampling.

The **exposure-response layer** was fitted in a separate cohort of 118 female
BALB/c mice with chronic-stage *Trypanosoma cruzi* CL Brener infection
(1000 bioluminescent blood trypomastigotes; chronic stage reached 50-70 days
post-infection). These mice weighed 20.5-28.1 g (median 25 g) and received 10,
20, 30, 50 or 100 mg/kg per administration for 5, 10 or 20 days across ten
regimens (Table 1 of the source). Parasitological cure -- absence of
bioluminescence on both in vivo and ex vivo imaging after cyclophosphamide
immunosuppression -- was achieved in 83 of 118 mice.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Assmus_2025_benznidazole_mouse")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Assmus_2025_benznidazole_mouse.R`. The
table below collects them in one place for review. "S1 Code" is the NONMEM
control stream of the final PK model, published as supporting information.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` | `log(2.18)` 1/h | Table 2: K_A = 2.18 (RSE 15.9%, 95% CI 1.65-3.03); S1 Code `$THETA 3` |
| `lcl` | `log(0.0106)` L/h | Table 2: CL/F = 10.6 mL/h (RSE 5.6%, 95% CI 9.62-11.88); S1 Code `$THETA 1` |
| `lvc` | `log(0.0197)` L | Table 2: V_C/F = 19.7 mL (RSE 8.2%, 95% CI 17.05-23.18); S1 Code `$THETA 2` |
| `lfdepot` | `fixed(log(1))` | Methods: "Relative bioavailability (F, fixed to unity for the population)"; S1 Code `$THETA 4` = "1 FIX" |
| `e_wt_cl` | `fixed(0.75)` | Methods: allometric function on clearance, exponent 0.75, a priori; S1 Code `TVCL = THETA(1)*((WT/0.0194)**0.75)` |
| `e_wt_vc` | `fixed(1.0)` | Methods: allometric function on volume, exponent 1, a priori; S1 Code `TVV2 = THETA(2)*((WT/0.0194)**1.00)` |
| `e_dose_ka` | `-0.775` | Table 2: theta_Dose = -0.775 (RSE 13.3%, 95% CI -0.977 to -0.583); S1 Code `$THETA 5` |
| `etalcl` | `~ 0.0298` | S1 Code `$OMEGA` IIV_CL = 0.0298; Table 2 reports the back-transform 17.4 %CV |
| `etalvc` | `~ 0.0271` | S1 Code `$OMEGA` IIV_V2 = 0.0271; Table 2 reports the back-transform 16.6 %CV |
| `expSd` | `sqrt(0.122)` | Table 2: sigma = 0.122 (95% CI 0.087-0.188); S1 Code `$SIGMA 0.122` (a variance) |
| `mic` | `fixed(6.427)` ug/mL | Table 1 footnote: "T>IC90, Time above IC90 in plasma (6.427 ug/mL...)"; Methods: IC90,plasma = 24.7 uM; derivation in S1 Text |
| `logite0_cure_auc` | `-2.170` | Table 4, AUCinf column: Intercept (SE 0.566) |
| `e_auc_cure` | `0.0033` | Table 4, AUCinf column: LogOdds (SE 0.0006) |
| `logite0_cure_tmic` | `-1.859` | Table 4, T>IC90 column: Intercept (SE 0.521) |
| `e_tmic_cure` | `1.497` | Table 4, T>IC90 column: LogOdds (SE 0.289) |
| Reference weight 0.0194 kg | n/a | Methods: allometry "centered on the median weight of mice in the satellite PK study (19.4 g)"; S1 Code `(WT/0.0194)` |
| Reference dose 30 mg/kg | n/a | Methods: "individual doses centered on the median dose (30 mg/kg)"; S1 Code `(DOSE_PER_KG/30)` |
| `d/dt(depot)`, `d/dt(central)` | n/a | Results: one-compartment disposition with first-order absorption; S1 Code `ADVAN5 TRANS1`, `COMP(1)` -> `COMP(2)`, `K12 = KA`, `K20 = CL/V2` |
| `ka <- exp(lka) * (DOSE_BZN_MGKG/30)^e_dose_ka` | n/a | Methods (power function on dose, centered on the median); S1 Code `COV1 = (DOSE_PER_KG/30)**THETA(5)`, `TVKA = THETA(3)*COV1` |
| `Cc ~ lnorm(expSd)` | n/a | Methods: "additive error on the log-transformed observed concentrations"; S1 Code `$ERROR Y = LOG(IPRED) + EPS(1)` |
| `d/dt(auc_total) <- Cc` | n/a | Methods, "Simulation of benznidazole exposure": cumulative AUC to infinity extracted from simulated profiles |
| `d/dt(t_above_mic) <- (Cc >= mic)/24` | n/a | Methods: "the time above the target concentration in mice (T>IC90) ... was calculated"; reported in days (Table 1) |
| `pcure_* <- 1/(1 + exp(-(logite0_* + e_*)))` | n/a | Equation 1: `logit(p) = beta0 + beta1*x` |

```{r model}
mod <- readModelDb("Assmus_2025_benznidazole_mouse")
```

# Part 1 -- Population PK in satellite mice

## Virtual cohort

Original observed data are not publicly available. The satellite cohort below
reproduces the published design: three single-dose groups, weights drawn to
match the reported 18.1-21.6 g range with a 19.4 g median and 0.92 g SD. The
physical dose is computed per mouse from its own weight, exactly as the study
did ("Doses, individually adjusted according to each mouse's weight").

```{r cohort-pk}
set.seed(20250513)

n_per_arm <- 200L  # cap is 200 participants per arm

make_pk_arm <- function(mgkg, n, id_offset) {
  wt <- pmin(pmax(rnorm(n, mean = 0.0194, sd = 0.00092), 0.0181), 0.0216)
  subj <- tibble(
    id            = id_offset + seq_len(n),
    WT            = wt,
    DOSE_BZN_MGKG = mgkg,
    arm           = paste0(mgkg, " mg/kg")
  )
  doses <- subj |>
    mutate(time = 0, evid = 1L, amt = DOSE_BZN_MGKG * WT, cmt = "depot")
  obs <- subj |>
    tidyr::crossing(time = seq(0, 24, by = 0.25)) |>
    mutate(evid = 0L, amt = NA_real_, cmt = "central")
  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}

events_pk <- bind_rows(
  make_pk_arm(10,  n_per_arm, id_offset = 0L),
  make_pk_arm(30,  n_per_arm, id_offset = 1000L),
  make_pk_arm(100, n_per_arm, id_offset = 2000L)
)
# No (id, time, evid) triple may repeat: `unique()` must NOT be applied first,
# or the check is vacuous (deduplicating before asking for duplicates).
stopifnot(!anyDuplicated(events_pk[, c("id", "time", "evid")]))
```

## Simulation

```{r simulate-pk}
sim_pk <- rxode2::rxSolve(
  mod,
  events = events_pk,
  keep   = c("arm", "WT", "DOSE_BZN_MGKG")
) |>
  as.data.frame()

stopifnot(dplyr::n_distinct(sim_pk$id) == 3L * n_per_arm)
stopifnot(all(sim_pk$Cc >= 0))
```

## Replicating Figure 1 (visual predictive check)

```{r figure-1}
# Replicates Figure 1 of Assmus 2025: simulated 5th / 50th / 95th percentiles
# of benznidazole plasma concentration by dose group, with the 5 ng/mL LLOQ.
sim_pk |>
  filter(!is.na(Cc), time > 0) |>
  mutate(arm = factor(arm, levels = c("10 mg/kg", "30 mg/kg", "100 mg/kg"))) |>
  group_by(arm, time) |>
  summarise(
    Q05 = quantile(sim, 0.05),
    Q50 = quantile(sim, 0.50),
    Q95 = quantile(sim, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "red") +
  geom_hline(yintercept = 0.005, linetype = "dashed") +
  facet_wrap(~arm) +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)", y = "Benznidazole (ug/mL)",
    title = "Figure 1 - simulated concentration-time profiles by dose group",
    caption = paste(
      "Replicates Figure 1 of Assmus 2025. Red = median, band = 5th-95th",
      "percentile of the simulated observations (residual error included).",
      "Dashed line = LLOQ (0.005 ug/mL)."
    )
  )
```

The dose-dependent absorption is visible directly: the peak shifts later and
broadens as the dose rises, because `ka` falls from 5.11 to 0.86 1/h between
10 and 100 mg/kg.

```{r ka-by-dose}
tibble(`Dose (mg/kg)` = c(10, 30, 100)) |>
  mutate(
    `KA, model (1/h)`     = round(2.18 * (`Dose (mg/kg)` / 30)^(-0.775), 2),
    `KA, published (1/h)` = c(5.11, 2.18, 0.86)
  ) |>
  knitr::kable(caption = paste(
    "Absorption rate constant by dose level. Published values from the",
    "Results section, 'Pharmacokinetics of benznidazole'."
  ))
```

## PKNCA validation

```{r pknca}
sim_nca <- sim_pk |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, arm)

# Guarantee a time = 0 row per (id, arm); pre-dose Cc = 0 for oral dosing.
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, arm) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, arm, time, .keep_all = TRUE) |>
  arrange(id, arm, time)

stopifnot(nrow(sim_nca) > 0, all(sim_nca$Cc >= 0))

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

dose_df <- events_pk |>
  filter(evid == 1L) |>
  select(id, time, amt, arm)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)

intervals <- data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

### Comparison against the published secondary PK parameters

S2 Table of the source reports model-derived secondary parameters as
median (5th-95th percentile) per dose group.

```{r nca-compare}
published <- tibble::tribble(
  ~arm,        ~cmax, ~tmax, ~aucinf.obs, ~half.life,
  "10 mg/kg",    8.0,  0.49,        18.1,       1.23,
  "30 mg/kg",   19.0,  0.86,        57.3,       1.33,
  "100 mg/kg",  44.7,  1.47,       186.0,       1.39
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "arm",
  units         = c(cmax = "ug/mL", aucinf.obs = "ug*h/mL",
                    tmax = "h", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated vs. published secondary PK parameters (S2 Table of Assmus 2025;",
    "medians). * differs from reference by >20%."
  ),
  align = c("l", "l", "r", "r", "r")
)
```

Every parameter agrees with S2 Table well inside the 20% tolerance. The
residual few-percent gaps on Cmax and AUCinf reflect the fact that the paper's
medians come from the 52 real mice at their individually measured weights and
their *actual* administered doses ("The exact dose administered to each mouse
was calculated based on the measured concentration of the formulation and the
animal's body weight"), whereas this cohort uses nominal mg/kg dose levels.

# Part 2 -- Simulated exposure in the efficacy regimens

Table 1 of the source reports median exposure metrics for the ten efficacy
regimens, simulated from this PK model at the 25 g median weight of the
efficacy cohort. Because those are typical-value simulations, the random
effects are zeroed here.

```{r regimens}
mod_typical <- rxode2::zeroRe(mod)

wt_eff <- 0.025  # 25 g, median weight of mice in the efficacy studies

regimens <- tibble::tribble(
  ~regimen, ~label,                     ~mgkg, ~n_dose, ~ii,
  "a",      "100 mg/kg, 10 days, QD",     100,      10,  24,
  "b",      "100 mg/kg, 5 days, QD",      100,       5,  24,
  "c",      "50 mg/kg, 10 days, BID",      50,      20,  12,
  "d",      "50 mg/kg, 10 days, QD",       50,      10,  24,
  "e",      "30 mg/kg, 20 days, QD",       30,      20,  24,
  "f",      "30 mg/kg, 10 days, QD",       30,      10,  24,
  "g",      "30 mg/kg, 5 days, QD",        30,       5,  24,
  "h",      "20 mg/kg, 10 days, QD",       20,      10,  24,
  "i",      "10 mg/kg, 20 days, QD",       10,      20,  24,
  "j",      "10 mg/kg, 10 days, QD",       10,      10,  24
)

simulate_regimen <- function(mgkg, n_dose, ii) {
  last_dose <- (n_dose - 1) * ii
  ev <- rxode2::et(amt = mgkg * wt_eff, ii = ii, until = last_dose, cmt = "depot") |>
    rxode2::et(seq(0, last_dose + 120, by = 0.02), cmt = "central")
  d <- as.data.frame(ev)
  d$WT <- wt_eff
  d$DOSE_BZN_MGKG <- mgkg
  rxode2::rxSolve(mod_typical, d, omega = NA, returnType = "data.frame",
                  atol = 1e-10, rtol = 1e-10)
}

sims <- lapply(seq_len(nrow(regimens)), function(i) {
  s <- simulate_regimen(regimens$mgkg[i], regimens$n_dose[i], regimens$ii[i])
  s$regimen <- regimens$regimen[i]
  s$label   <- regimens$label[i]
  s
})
sim_reg <- bind_rows(sims) |> filter(!is.na(Cc))
stopifnot(dplyr::n_distinct(sim_reg$regimen) == nrow(regimens))
```

## Replicating Figure 2 (typical profiles by regimen)

```{r figure-2, fig.height = 7}
# Replicates Figure 2 of Assmus 2025: median simulated benznidazole profiles
# for each efficacy regimen in a 25 g mouse, against the plasma IC90 and its
# 2-fold bracket.
ic90 <- 6.427

sim_reg |>
  filter(time <= 24 * 21) |>
  mutate(label = factor(label, levels = regimens$label)) |>
  ggplot(aes(time / 24, Cc)) +
  geom_line() +
  geom_hline(yintercept = ic90, colour = "red") +
  geom_hline(yintercept = c(ic90 / 2, ic90 * 2), colour = "steelblue",
             linetype = "dashed") +
  facet_wrap(~label, ncol = 2) +
  labs(
    x = "Time (days)", y = "Benznidazole (ug/mL)",
    title = "Figure 2 - typical profiles for the ten efficacy regimens",
    caption = paste(
      "Replicates Figure 2 of Assmus 2025. Red = IC90,plasma (6.427 ug/mL);",
      "dashed = 2-fold higher and lower target concentrations."
    )
  )
```

## Replicating Table 1 (exposure metrics)

`AUC12` and `AUC24` are read directly off the model's cumulative-AUC state at
12 h and 24 h; `AUCinf` and `T>IC90` are the terminal values of the
`auc_total` and `t_above_mic` states.

```{r table-1}
at_time <- function(df, tt, col) {
  df[[col]][which.min(abs(df$time - tt))]
}

exposure <- bind_rows(lapply(sims, function(s) {
  tibble(
    regimen = s$regimen[1],
    cmax    = max(s$Cc, na.rm = TRUE),
    auc12   = at_time(s, 12, "auc_total"),
    auc24   = at_time(s, 24, "auc_total"),
    aucinf  = max(s$auc_total),
    tmic    = max(s$t_above_mic),
    p_auc   = max(s$pcure_auc),
    p_tmic  = max(s$pcure_tmic)
  )
}))

published_t1 <- tibble::tribble(
  ~regimen, ~cum_dose, ~r_cmax, ~r_auc12, ~r_auc24, ~r_aucinf, ~r_tmic, ~n_cured, ~n_total,
  "a", 1000, 46.00, 193.4, 194.5, 1945, 2.86, 25, 27,
  "b",  500, 46.08, 195.4, 196.9,  985, 1.44, 17, 17,
  "c", 1000, 28.25,  96.8, 194.1, 1946, 3.95,  6,  6,
  "d",  500, 28.11,  96.9,  97.2,  972, 1.96,  2,  3,
  "e",  600, 19.07,  58.3,  58.6, 1173, 2.85, 23, 29,
  "f",  300, 18.97,  58.1,  58.3,  583, 1.41,  9, 11,
  "g",  150, 19.01,  58.8,  59.1,  295, 0.72,  0,  9,
  "h",  200, 13.70,  38.8,  38.9,  389, 1.01,  0,  4,
  "i",  200,  7.67,  19.5,  19.6,  391, 0.68,  1,  6,
  "j",  100,  7.65,  19.4,  19.4,  194, 0.33,  0,  6
)

tab1 <- exposure |>
  left_join(published_t1, by = "regimen") |>
  left_join(regimens |> select(regimen, label), by = "regimen") |>
  mutate(
    d_cmax   = 100 * (cmax   - r_cmax)   / r_cmax,
    d_auc12  = 100 * (auc12  - r_auc12)  / r_auc12,
    d_auc24  = 100 * (auc24  - r_auc24)  / r_auc24,
    d_aucinf = 100 * (aucinf - r_aucinf) / r_aucinf,
    d_tmic   = 100 * (tmic   - r_tmic)   / r_tmic
  )

tab1 |>
  transmute(
    Regimen                = label,
    `Cmax sim`             = round(cmax, 2),
    `Cmax pub`             = r_cmax,
    `AUC12 sim`            = round(auc12, 1),
    `AUC12 pub`            = r_auc12,
    `AUC24 sim`            = round(auc24, 1),
    `AUC24 pub`            = r_auc24,
    `AUCinf sim`           = round(aucinf, 0),
    `AUCinf pub`           = r_aucinf,
    `T>IC90 sim (d)`       = round(tmic, 2),
    `T>IC90 pub (d)`       = r_tmic
  ) |>
  knitr::kable(caption = paste(
    "Replicates Table 1 of Assmus 2025: simulated median exposures for a 25 g",
    "mouse. Cmax in ug/mL, AUC in ug*h/mL, T>IC90 in days."
  ))
```

```{r table-1-check}
worst <- tab1 |>
  summarise(
    Cmax   = max(abs(d_cmax)),
    AUC12  = max(abs(d_auc12)),
    AUC24  = max(abs(d_auc24)),
    AUCinf = max(abs(d_aucinf)),
    `T>IC90` = max(abs(d_tmic))
  )
knitr::kable(round(worst, 2), caption = "Worst-case absolute % difference vs. Table 1.")

# Every metric must reproduce Table 1 to within 3%. This is a real gate: the
# published values were generated from n = 10 stochastic replicates, so a few
# tenths of a percent of Monte-Carlo noise is expected, but nothing more.
stopifnot(
  nrow(tab1) == 10L,
  max(abs(tab1$d_cmax))   < 3,
  max(abs(tab1$d_auc12))  < 3,
  max(abs(tab1$d_auc24))  < 3,
  max(abs(tab1$d_aucinf)) < 3,
  max(abs(tab1$d_tmic))   < 3
)
```

All five exposure metrics for all ten regimens reproduce Table 1 to within 3%.
Note the structural signature of the BID regimen (c): its `AUC12` is half that
of the matched 100 mg/kg QD regimen (a) while its `AUC24` and `AUCinf` are the
same, and its `T>IC90` is the longest of any regimen (3.95 days) -- splitting
the daily dose buys time above target without changing total exposure.

# Part 3 -- Exposure-response for parasitological cure

## Predicted cure probability by regimen

The two logistic models carried inside the model object are evaluated at each
regimen's terminal exposure and compared against the observed cure fraction
reported in Table 1.

```{r exposure-response}
er <- tab1 |>
  mutate(
    obs_rate = n_cured / n_total,
    lo = mapply(function(x, n) stats::binom.test(x, n)$conf.int[1], n_cured, n_total),
    hi = mapply(function(x, n) stats::binom.test(x, n)$conf.int[2], n_cured, n_total)
  )

er |>
  transmute(
    Regimen                     = label,
    `AUCinf (ug*h/mL)`          = round(aucinf, 0),
    `P(cure) | AUCinf`          = round(p_auc, 3),
    `T>IC90 (days)`             = round(tmic, 2),
    `P(cure) | T>IC90`          = round(p_tmic, 3),
    `Observed cured / total`    = paste0(n_cured, "/", n_total),
    `Observed rate`             = round(obs_rate, 3)
  ) |>
  knitr::kable(caption = paste(
    "Model-predicted probability of sterile parasitological cure vs. the",
    "observed cure rate in each regimen (Table 1 of Assmus 2025)."
  ))
```

```{r figure-3}
# Replicates the lower panel of Figure 3 of Assmus 2025: predicted probability
# of cure across the exposure range, with the observed per-regimen cure rates.
auc_grid <- tibble(
  aucinf = seq(0, 2200, length.out = 200)
) |>
  mutate(p = 1 / (1 + exp(-(-2.170 + 0.0033 * aucinf))))

ggplot(auc_grid, aes(aucinf, p)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_pointrange(
    data = er,
    aes(x = aucinf, y = obs_rate, ymin = lo, ymax = hi),
    inherit.aes = FALSE
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Cumulative AUCinf (ug*h/mL)", y = "Probability of parasitological cure",
    title = "Figure 3d - exposure-response for sterile cure",
    caption = paste(
      "Replicates the AUCinf panel of Figure 3 of Assmus 2025. Line = logistic",
      "regression (Table 4); points = observed cure rate per regimen with an",
      "exact binomial 95% CI."
    )
  )
```

## Replicating Table 5 (exposure required for a target cure probability)

Inverting each logistic model, `x = (logit(p) - beta0) / beta1`.

```{r table-5}
invert <- function(p, b0, b1) (log(p / (1 - p)) - b0) / b1

targets <- c(0.50, 0.90, 0.95, 0.99)

tab5 <- tibble(
  `Predicted probability of cure` = paste0(targets * 100, "%"),
  `AUCinf sim (ug*h/mL)`  = round(invert(targets, -2.170, 0.0033), 0),
  `AUCinf pub (ug*h/mL)`  = c(656, 1319, 1545, 2044),
  `T>IC90 sim (days)`     = round(invert(targets, -1.859, 1.497), 2),
  `T>IC90 pub (days)`     = c(1.24, 2.71, 3.21, 4.31)
)

knitr::kable(tab5, caption = paste(
  "Replicates Table 5 of Assmus 2025: benznidazole exposure in plasma required",
  "for a given predicted probability of parasitological cure in mice."
))

# Table 5 is an independent answer key for both logistic coefficient pairs:
# it was computed by the authors from the same Table 4 estimates, so an exact
# match confirms the intercepts and slopes were transcribed correctly.
stopifnot(
  max(abs(tab5$`AUCinf sim (ug*h/mL)` - tab5$`AUCinf pub (ug*h/mL)`) /
        tab5$`AUCinf pub (ug*h/mL)`) < 0.01,
  max(abs(tab5$`T>IC90 sim (days)` - tab5$`T>IC90 pub (days)`) /
        tab5$`T>IC90 pub (days)`) < 0.01
)
```

Both inverted models land within 1% of the published thresholds, confirming the
transcribed intercepts and slopes.

## The other four exposure metrics

Table 4 of the source reports **six** binary univariate logistic regressions:
the five simulated plasma-exposure metrics named in the Methods (`CMAX`,
`AUC12`, `AUC24`, `AUCinf`, `T>IC90`) plus cumulative dose. Two of the six are
ODE-native and are carried inside the model object (`AUCinf` via `auc_total`,
`T>IC90` via `t_above_mic`, above). The remaining four are summary quantities
with no natural ODE state, and are reproduced here from the same Table 4
coefficients applied to the exposures simulated above, so that all six
regressions appear in this vignette.

```{r other-metrics}
other <- tibble::tribble(
  ~metric,            ~b0,     ~b1,
  "Cumulative dose",  -2.157,  0.006,
  "CMAX",             -2.312,  0.137,
  "AUC12",            -1.394,  0.028,
  "AUC24",            -1.375,  0.026
)

# Cumulative dose is a design quantity, not a simulated one: dose per
# administration times the number of administrations (Table 1, column 1).
cum_dose_sim <- regimens$mgkg * regimens$n_dose
stopifnot(identical(regimens$regimen, tab1$regimen),
          all(cum_dose_sim == tab1$cum_dose))

bind_rows(lapply(seq_len(nrow(other)), function(i) {
  metric <- other$metric[i]
  x <- switch(
    metric,
    `Cumulative dose` = cum_dose_sim,
    CMAX  = tab1$cmax,
    AUC12 = tab1$auc12,
    AUC24 = tab1$auc24
  )
  tibble(
    Regimen  = tab1$label,
    Metric   = metric,
    exposure = round(x, 1),
    pcure    = round(1 / (1 + exp(-(other$b0[i] + other$b1[i] * x))), 3)
  )
})) |>
  tidyr::pivot_wider(names_from = Metric,
                     values_from = c(exposure, pcure)) |>
  knitr::kable(caption = paste(
    "Predicted cure probability from the cumulative-dose, CMAX, AUC12 and",
    "AUC24 univariate logistic models (Table 4 of Assmus 2025)."
  ))
```

Cumulative dose and `AUCinf` are perfectly collinear in this design
(Table 3: `R^2 = 1.00`), and Table 4 accordingly reports near-identical
diagnostics for the two (AIC 99.1 vs. 98.8; identical accuracy, sensitivity,
specificity and MCC). The paper focuses on `AUCinf` because a plasma-exposure
metric, unlike an administered dose, carries translational value for human
dose prediction.

The paper's own conclusion is that these metrics cannot be told apart:
`AUCinf` has the best goodness of fit (AIC 98.8 vs. 105.1-108.5) but every
delta-AIC is under 10, and the areas under the ROC curve for `CMAX`, `AUCinf`
and `T>IC90` all exceed 85% with overlapping 95% CIs. That collinearity
(Table 3: `R^2 > 0.95` within each of two clusters) is the paper's central
caveat and the reason it calls for dose-fractionation studies.

# Assumptions and deviations

- **Volume and clearance units.** The paper reports `V_C/F = 19.7 mL` and
  `CL/F = 10.6 mL/h`. The model file carries these as `0.0197 L` and
  `0.0106 L/h` so that a dose in mg yields a concentration in mg/L, which is
  numerically identical to the ug/mL used throughout the paper. No values were
  changed; only the unit prefix.
- **`sigma` is a variance, not an SD.** Table 2 labels the residual-error row
  `sigma` and gives 0.122, but the S1 Code `$SIGMA` block -- whose `$THETA` and
  `$OMEGA` values match Table 2 exactly -- shows this is the NONMEM variance.
  The model therefore uses `expSd = sqrt(0.122) = 0.349` as the log-scale
  residual SD. The same reading is what makes Table 2's IIV column internally
  consistent: `$OMEGA` 0.0298 and 0.0271 are variances that back-transform to
  the reported 17.4 %CV and 16.6 %CV.
- **IIV on KA and F is omitted rather than written as `fixed(0)`.** The paper
  estimated both at under 10 %CV and fixed them to zero in the final model
  (S1 Code `$OMEGA` entries 3 and 4 are `0 FIX`). Encoding a zero-variance
  diagonal would make OMEGA singular and break rxode2's Cholesky sampler at
  simulation time; a fixed-zero variance is mathematically identical to having
  no eta at all.
- **Only two of the six exposure-response models are inside the model
  object.** `AUCinf` and `T>IC90` are cumulative quantities that an ODE can
  carry, so they are encoded as the `auc_total` and `t_above_mic` states with
  their logistic layers. Cumulative dose, `CMAX`, `AUC12` and `AUC24` are
  design or summary NCA quantities with no natural ODE state (a running
  maximum in particular cannot be integrated without a stiff pseudo-state that
  would destabilise the solver), so their Table 4 coefficients are applied in
  this vignette instead. No endpoint was dropped: all six regressions reported
  in Table 4 appear above.
- **Table 1 and the Results prose disagree on the size of regimen (f).**
  Table 1 reports 9/11 cured (81.8%) for 30 mg/kg QD x 10 days, while the
  Results text says "cure rates reached approximately 80% in mice treated for
  10 days (9/10; scenario f)". Table 1 is the value used here, because its
  denominators are the internally consistent set: they sum to the 118 mice
  stated throughout the paper (the prose's 9/10 would give 117), and its
  numerators sum to the stated 83 cured. No correction notice was found for
  this article.
- **The logistic layer is a regimen-level, not a per-mouse, relationship.** The
  authors regressed per-mouse binary cure status on the *median simulated*
  exposure for that mouse's regimen -- one distinct predictor value per
  regimen, not per animal. Evaluating `pcure_auc` / `pcure_tmic` on an
  individual simulated mouse (i.e. with IIV switched on) therefore extrapolates
  the relationship beyond how it was fitted. Use `zeroRe()` for typical-value
  work, as this vignette does.
- **The PK and efficacy cohorts are different animals.** The PK layer was
  estimated in 52 *uninfected* satellite mice (median 19.4 g); the
  exposure-response layer was fitted in 118 *chronically infected* mice (median
  25 g). The model assumes benznidazole PK is unchanged by infection, which the
  paper assumes implicitly and does not test.
- **Nominal rather than measured doses.** The satellite cohort here doses at
  nominal mg/kg levels. The study computed each mouse's actual dose from the
  measured formulation concentration and the animal's weight, which is the most
  likely reason simulated Cmax and AUCinf sit a few percent below the S2 Table
  medians.
- **Body weights are simulated.** Individual weights are not published, so the
  cohort draws from a truncated normal matched to the reported median (19.4 g),
  SD (0.92 g) and range (18.1-21.6 g).
- **`IC90,plasma` is held fixed at the in vitro value.** The paper's own
  sensitivity analysis (S4 and S5 Tables) sweeps it across a 10-fold range and
  finds `T>IC90` highly sensitive to the choice. Override it with
  `rxode2::rxSolve(mod, ..., params = c(mic = <value>))` to reproduce that
  analysis.
- **The PLS-DA multivariate analysis is not extracted.** It is reported as
  giving no improvement over the univariate models, and the paper publishes
  only VIP scores (1.47, 1.40, 1.37 for `AUCinf`, `T>IC90`, `CMAX`) rather
  than a usable set of latent-variable coefficients.
