---
title: "Selexipag pediatric dose selection and confirmation (Axelsen 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Selexipag pediatric dose selection and confirmation (Axelsen 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE}
.meta_adult <- rxode2::rxode(readModelDb("Axelsen_2024_selexipag_adult"))
.meta_ped   <- rxode2::rxode(readModelDb("Axelsen_2024_selexipag_pediatric"))
```

* Citation: `r .meta_adult$meta$reference`
* Article: <https://doi.org/10.1002/psp4.13231>
* PMCID: [PMC11646929](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC11646929/)

Axelsen 2024 does two things. First, it takes the adult GRIPHON
population PK model for selexipag and its active metabolite
JNJ-68006861 (also known as ACT-333679) -- originally fitted in Monolix
and published as Krause 2017, and available in this library as
`Krause_2017_selexipag` -- re-estimates it in NONMEM, and uses it to
extrapolate a body-weight-banded pediatric dose regimen. Second, it
tests that regimen in 63 pediatric patients with pulmonary arterial
hypertension (PAH) aged 2 to <18 years (NCT03492177) and updates the
model with the pediatric data by NONMEM `BAYES` estimation, using the
adult estimates as priors.

The paper therefore reports **two parameter sets for one structural
model** (Table 1, "Adult population PK model" and "Pediatric population
PK model" columns). Following the library's replicate-the-author's-
structure policy, each is packaged as its own model file:

| Model file | Table 1 column | Fitted to |
|---|---|---|
| `Axelsen_2024_selexipag_adult` | Adult | Adult GRIPHON data (NCT01106014), re-estimated in NONMEM |
| `Axelsen_2024_selexipag_pediatric` | Pediatric | NCT03492177 pediatric data, Bayesian update of the adult model |

Both are validated in this single vignette.

## Structure

Dose enters a `depot` after an absorption lag and is absorbed
first-order (`ka`) into a two-compartment selexipag disposition
(`central` / `peripheral1`, exchange `kx12` / `kx21`). Selexipag leaves
`central` by **two parallel routes**: a linear apparent clearance
`CL/F` and the first-order metabolite-formation rate constant `kmet`,
which transfers drug into the metabolite's `central_act`. The
metabolite has its own two-compartment disposition (`kx34` / `kx43`)
and first-order elimination `km`.

This "`CL` excludes `kmet`" convention is the single most important
structural detail in the paper, and it is stated three independent
ways:

1. Table 1 describes `CL` as "Apparent clearance of parent (**not
   including kmet**)".
2. The printed exposure formula uses `Vp * kmet + CL/F` as the
   selexipag denominator (see the closed-form gate below).
3. The Supplementary Table S3 NONMEM control stream's `ADVAN5` transfer
   block sets `k2T0 = CL/Vp` (elimination) and `k2T4 = kmet`
   (parent -> metabolite) as separate outflows from the parent central
   compartment.

Total apparent selexipag clearance is therefore `CL/F + Vp/F * kmet`,
and the fraction metabolised is `Vp*kmet / (Vp*kmet + CL/F)` ~ 0.38.

```{r model-load}
mod_adult <- rxode2::rxode(readModelDb("Axelsen_2024_selexipag_adult"))
mod_ped   <- rxode2::rxode(readModelDb("Axelsen_2024_selexipag_pediatric"))
```

# Population

## Adult (GRIPHON, NCT01106014)

`r .meta_adult$meta$population$disease_state` The body-weight-exposure
relationship was identified over
`r .meta_adult$meta$population$weight_range` Adult exposures used as
the pediatric comparator come from N = `r .meta_adult$meta$population$n_subjects`
adults (Axelsen 2024 Table 3).

## Pediatric (NCT03492177)

`r .meta_ped$meta$population$disease_state`

* **N** = `r .meta_ped$meta$population$n_subjects` participants with full PK profiles (+3 with troughs only; 63 enrolled).
* **Age**: `r .meta_ped$meta$population$age_range`
* **Weight**: `r .meta_ped$meta$population$weight_range`
* **Sex**: `r .meta_ped$meta$population$sex_female_pct`
* **Race**: `r .meta_ped$meta$population$race_ethnicity`
* **Dosing**: `r .meta_ped$meta$population$dose_range`
* **Co-medication**: `r .meta_ped$meta$population$co_medication`

# Source trace

Every `ini()` value and every non-trivial `model()` equation, with the
location it was read from. "Table S3 control stream" refers to the
NONMEM model code distributed as Supplementary Table S3 of the article.

| Model element | Value (adult / pediatric) | Source |
|---|---|---|
| `lka` | 0.688 / 0.663 1/h | Table 1, `ka`; control stream `$THETA 1` |
| `lkmet` | 0.887 / 0.868 1/h | Table 1, `kmet`; `$THETA 2` |
| `lcl` | 18 / 18.9 L/h | Table 1, `CL`; `$THETA 3` |
| `lvc` | 12.2 / 12.4 L | Table 1, `Vp`; `$THETA 4` |
| `lk12` | 0.101 / 0.115 1/h | Table 1, `kx12`; `$THETA 5` |
| `lk21` | 0.0521 / 0.0563 1/h | Table 1, `kx21`; `$THETA 6` |
| `lkm_act` | 0.468 / 0.445 1/h | Table 1, `km`; `$THETA 7` |
| `lk34_act` | 0.898 / 0.83 1/h | Table 1, `kx34`; `$THETA 8` |
| `lk43_act` | 0.183 / 0.168 1/h | Table 1, `kx43`; `$THETA 9` |
| `lvc_act` | 5.88 / 6.05 L | Table 1, `Vm`; `$THETA 10` |
| `logittlag` | -0.69 (FIX, both) | Control stream `$THETA 11` `-0.69 FIX ; log(Tlag1Half/(1-Tlag1Half)) (0.334)`; Table 1 `Tlag1Half` = 0.334 h (FIX) |
| `e_wt_cl` | 0.546 / 0.828 | Table 1, `beta_CL(WEIGHTBL)`; `$THETA 12` |
| `e_tbili_cl` | -0.442 / -0.363 | Table 1, `beta_CL(BILIBL)`; `$THETA 13` |
| `e_wt_vc_act` | 0.803 / 0.542 | Table 1, `beta_Vm(WEIGHTBL)`; `$THETA 14` |
| `e_wt_vc` | 1.04 / 0.805 | Table 1, `beta_Vp(WEIGHTBL)`; `$THETA 15` |
| `e_sexf_km_act` | 0.148 / 0.145 | Table 1, `beta_km(SEXN_0)`; `$THETA 16` |
| `e_era_km_act` | 0.164 / 0.186 | Table 1, `beta_km(COPAH1_1)`; `$THETA 17` |
| `e_pde5_km_act` | 0.0633 / 0.0495 | Table 1, `beta_km(COPAH2_1)`; `$THETA 18` |
| `e_combo_km_act` | 0.358 / 0.368 | Table 1, `beta_km(COPAH3_1)`; `$THETA 19` |
| All `eta*` variances | Table 1 `omega(...)` squared | Table 1 note "omega values reported as standard deviation"; control stream `$OMEGA` / `$OMEGAP` carry the `STANDARD` keyword |
| `propSd`, `propSd_act` | 0.754, 0.489 / 0.694, 0.455 | Table 1 `error_PROP1` / `error_PROP2`; control stream `$SIGMA` holds their squares (0.567954 = 0.753627^2, 0.239527 = 0.489415^2), confirming the printed values are SDs |
| ODE: `d/dt(central)` outflow `kel + kmet` | -- | Control stream `k2T0 = CL/Vp`, `k2T4 = kmet`; Table 1 "`CL` ... not including kmet"; printed AUC formula denominator `Vp*kmet + CL/F` |
| `alag(depot) <- 2 * expit(logittlag + eta)` | -- | Control stream `Tlag1Half = EXP(T)/(1+EXP(T))` and `ALAG1 = 2*Tlag1Half` |
| Covariate form: power (continuous), `exp(beta*I)` (categorical) | -- | Control stream builds each `MU_i` additively on the log scale then exponentiates (e.g. `MU_3 = THETA(3) + THETA(12)*log(WEIGHTBL/70) + THETA(13)*log(BILIBL/10)`; `CL = EXP(T_CL)`) |
| Male indicator `(1 - SEXF)` | -- | Control stream `SEXN_0 = 1 IF(SEXN.EQ.0)` with `; reference: 1`; Table 1 labels the coefficient "Gender male on km" |
| Reference values 70 kg, 10 umol/L | -- | Control stream `log(WEIGHTBL/70)`, `log(BILIBL/10)`; Table 1 "centered around 70 kg" / "centered around 10 umol/L" |
| Units: dose ug, conc ng/mL, time h | -- | Control stream header comment "Dose: ug / Concentration: ng/mL / Time: hours" |

# Validation 1 -- the paper's own closed-form exposure formula

Axelsen 2024 Methods prints the combined exposure metric used for the
whole dose-selection exercise:

$$
\mathrm{AUC}_{\tau,ss,combined}
= \tfrac{1}{38}\,\mathrm{AUC}_{\tau,ss,\mathrm{selexipag}}
+ \tfrac{37}{38}\,\mathrm{AUC}_{\tau,ss,\mathrm{JNJ}}
= \tfrac{1}{38}\,\mathrm{Dose}\left\{
\frac{1}{V_p k_{met} + CL/F}
+ 37\,\frac{V_p k_{met}}{V_p k_{met} + CL/F}\cdot\frac{1}{V_M k_m}
\right\}
$$

The 1/38 and 37/38 weights are the relative IP-receptor potencies: the
Introduction states the active metabolite is 37 times more potent than
selexipag.

This is an exact analytical consequence of the ODE system, so solving
the packaged model to steady state and integrating must reproduce it to
numerical precision. Because both sides of this comparison use the
*same* individual parameters, the difference is pure integration error
and a tight bound is the correct gate.

```{r closed-form-gate}
tau     <- 12    # h, b.i.d.
n_doses <- 28    # 14 days -- ample for steady state given kx21 = 0.052 1/h

# Steady-state event table. Observation rows are tagged with dvid = 1
# (the first endpoint); rxode2 returns BOTH Cc and Cc_act at every
# observation row, so the endpoint tag only marks the row as an
# observation. Compartments are never referenced by an observable name.
make_events <- function(cohort, obs_times) {
  dose <- tidyr::crossing(id = cohort$id, dose_idx = seq_len(n_doses)) |>
    dplyr::mutate(time = (dose_idx - 1) * tau, evid = 1L, cmt = "depot") |>
    dplyr::select(-dose_idx)
  obs <- tidyr::crossing(id = cohort$id, time = (n_doses - 1) * tau + obs_times) |>
    dplyr::mutate(evid = 0L, amt = NA_real_, dvid = 1L)
  dplyr::bind_rows(dose, obs) |>
    dplyr::left_join(cohort, by = "id") |>
    dplyr::mutate(amt = dplyr::if_else(evid == 1L, dose_ug, amt)) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

# Combined exposure, exactly as printed above.
auc_combined <- function(dose, vc, kmet, cl, vc_act, km_act) {
  tot <- vc * kmet + cl
  (1 / 38) * dose * (1 / tot + 37 * (vc * kmet / tot) / (vc_act * km_act))
}

# One typical patient per model: 70 kg, 10 umol/L bilirubin, female,
# on ERA + PDE5 inhibitor (the modal PAH-comedication stratum).
ref_subject <- tibble::tibble(
  id = 1L, dose_ug = 200, WT = 70, TBILI = 10, SEXF = 1,
  CONMED_ERA = 0, CONMED_PDE5I = 0, CONMED_ERA_PDE5I = 1
)

trap <- function(x, t) sum(diff(t) * (utils::head(x, -1) + utils::tail(x, -1)) / 2)

closed_form_check <- function(mod, label) {
  ev <- make_events(ref_subject, seq(0, tau, by = 0.02))
  s <- rxode2::rxSolve(rxode2::zeroRe(mod), ev, keep = "dose_ug",
                       returnType = "data.frame") |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::arrange(time)
  p <- s[1, ]
  tot <- p$vc * p$kmet + p$cl
  tibble::tibble(
    model = label,
    quantity = c("AUCtau,ss selexipag", "AUCtau,ss JNJ-68006861", "AUCtau,ss combined"),
    solved = c(trap(s$Cc, s$time),
               trap(s$Cc_act, s$time),
               (1 / 38) * trap(s$Cc, s$time) + (37 / 38) * trap(s$Cc_act, s$time)),
    closed_form = c(p$dose_ug / tot,
                    p$dose_ug * (p$vc * p$kmet / tot) / (p$vc_act * p$km_act),
                    auc_combined(p$dose_ug, p$vc, p$kmet, p$cl, p$vc_act, p$km_act))
  ) |>
    dplyr::mutate(pct_diff = 100 * (solved / closed_form - 1))
}

cf <- dplyr::bind_rows(
  closed_form_check(mod_adult, "Adult"),
  closed_form_check(mod_ped,   "Pediatric")
)

cf |>
  dplyr::rename("Model" = model, "Quantity" = quantity,
                "ODE solve" = solved, "Printed formula" = closed_form,
                "% diff" = pct_diff) |>
  knitr::kable(digits = 4,
               caption = "Steady-state exposure from the ODE solve vs. the closed form printed in Axelsen 2024 Methods.")

# Same parameters on both sides, so this is integration error only.
stopifnot(max(abs(cf$pct_diff)) < 0.5)
```

The packaged ODEs reproduce the paper's printed formula to better than
0.01%. This simultaneously confirms the compartment topology, the
`CL/F + Vp*kmet` total-clearance convention, and the units.

# Virtual pediatric cohort

Covariate distributions are taken from Supplementary Table S3
(descriptive statistics for the pediatric study covariates), by age
cohort. Continuous covariates are drawn log-normal with the published
mean and SD and truncated to the published range; sex and
PAH-comedication stratum are drawn from the published proportions.

Starting dose is **not** drawn -- it is derived from the protocol's
body-weight bands (Axelsen 2024 Figure 2 / Results): 100 ug b.i.d. for
>=9 to <25 kg, 150 ug for >=25 to <50 kg, 200 ug for >=50 kg. That the
resulting dose-group proportions reproduce the observed ones (Table S3)
is itself a check on the weight distributions.

```{r cohort}
n_per_arm <- 200   # per age cohort; library cap is 200 per arm

# Supplementary Table S3: mean (SD) [range] by age cohort.
cohort_spec <- tibble::tribble(
  ~cohort,        ~n_pub, ~wt_m, ~wt_sd, ~wt_lo, ~wt_hi, ~bili_m, ~bili_sd, ~bili_lo, ~bili_hi, ~p_female, ~p_none, ~p_era, ~p_pde5, ~p_both,
  "12-17 years",      21,  51.1,   14.3,     30,   93.5,    15.1,     16.8,        3,       84,     0.667,  0.0952,  0.143,   0.238,   0.524,
  "6-11 years",       21,  25.0,   6.49,   16.8,   36.4,    10.4,     6.77,        3,       32,     0.524,  0.0476, 0.0952,   0.571,   0.286,
  "2-5 years",        20,  16.4,   6.95,    9.9,   41.5,     7.9,     5.57,        3,       23,     0.500,  0.0000,  0.100,   0.150,   0.750
)

# Log-normal draw matched to a target arithmetic mean and SD, truncated
# to the published observed range by rejection.
rlnorm_mean_sd <- function(n, m, s, lo, hi) {
  sdlog <- sqrt(log1p((s / m)^2))
  mlog  <- log(m) - sdlog^2 / 2
  out <- numeric(0)
  while (length(out) < n) {
    x <- stats::rlnorm(2 * n, mlog, sdlog)
    out <- c(out, x[x >= lo & x <= hi])
  }
  out[seq_len(n)]
}

set.seed(20241213)
rxode2::rxSetSeed(20241213)

cohort <- purrr::pmap_dfr(cohort_spec, function(cohort, n_pub, wt_m, wt_sd, wt_lo, wt_hi,
                                                bili_m, bili_sd, bili_lo, bili_hi,
                                                p_female, p_none, p_era, p_pde5, p_both) {
  pah <- sample(c("none", "ERA", "PDE5", "ERA+PDE5"), n_per_arm, replace = TRUE,
                prob = c(p_none, p_era, p_pde5, p_both))
  tibble::tibble(
    cohort = cohort,
    WT     = rlnorm_mean_sd(n_per_arm, wt_m, wt_sd, wt_lo, wt_hi),
    TBILI  = rlnorm_mean_sd(n_per_arm, bili_m, bili_sd, bili_lo, bili_hi),
    SEXF   = stats::rbinom(n_per_arm, 1, p_female),
    CONMED_ERA       = as.integer(pah == "ERA"),
    CONMED_PDE5I     = as.integer(pah == "PDE5"),
    CONMED_ERA_PDE5I = as.integer(pah == "ERA+PDE5")
  )
}) |>
  dplyr::mutate(
    wt_group = dplyr::case_when(WT <  25 ~ ">=9 to <25 kg",
                                WT <  50 ~ ">=25 to <50 kg",
                                TRUE     ~ ">=50 kg"),
    dose_ug  = dplyr::case_when(WT <  25 ~ 100,
                                WT <  50 ~ 150,
                                TRUE     ~ 200),
    dose_group = paste0(dose_ug, " ug"),
    id = dplyr::row_number()
  )

cohort |>
  dplyr::count(cohort, dose_group) |>
  dplyr::group_by(cohort) |>
  dplyr::mutate(`Simulated %` = round(100 * n / sum(n), 1)) |>
  dplyr::ungroup() |>
  dplyr::select(-n) |>
  tidyr::pivot_wider(names_from = dose_group, values_from = `Simulated %`, values_fill = 0) |>
  dplyr::select("Age cohort" = cohort, "100 ug", "150 ug", "200 ug") |>
  dplyr::arrange(match(`Age cohort`, c("12-17 years", "6-11 years", "2-5 years"))) |>
  knitr::kable(caption = "Simulated starting-dose distribution from the body-weight bands. Observed (Table S3): 12-17 y = 0 / 42.9 / 57.1%; 6-11 y = 57.1 / 42.9 / 0%; 2-5 y = 95 / 5 / 0% for 100 / 150 / 200 ug.")
```

# Simulation

Sampling reproduces the study's own schedule -- predose and 1, 2, 4, 6,
8 and 12 h postdose (Methods, "Pediatric study design, participants,
and PK sampling") -- so the simulated NCA carries the same trapezoidal
and `tmax`-quantisation behaviour as the published NCA in Table 2.
Because the model is linear, dose-normalised exposure is
dose-independent, so simulating each participant at their starting dose
is equivalent to the paper's dose-normalisation of Week-1 or Week-12
profiles.

```{r simulate}
rxode2::rxSetSeed(20241213)
set.seed(20241213)

obs_times <- c(0, 1, 2, 4, 6, 8, 12)
keep_cols <- c("WT", "TBILI", "SEXF", "CONMED_ERA", "CONMED_PDE5I",
               "CONMED_ERA_PDE5I", "cohort", "wt_group", "dose_group", "dose_ug")

sim <- rxode2::rxSolve(
  mod_ped,
  make_events(cohort, obs_times),
  keep = keep_cols,
  returnType = "data.frame"
) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(time = time - (n_doses - 1) * tau)

# Per-subject individual parameters -> model-based AUCtau,ss,combined,
# which is exactly how Axelsen 2024 computed the values in Table 3.
ind <- sim |>
  dplyr::group_by(id) |>
  dplyr::slice(1) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    tot      = vc * kmet + cl,
    auc_p    = dose_ug / tot,                                    # AUCtau,ss selexipag
    auc_m    = dose_ug * (vc * kmet / tot) / (vc_act * km_act),  # AUCtau,ss JNJ-68006861
    auc_comb = auc_combined(dose_ug, vc, kmet, cl, vc_act, km_act)
  )
```

## Exact model-predicted exposure vs. Table 2

Before running NCA on a sampling grid, it is worth comparing the
**exact** model-predicted steady-state exposure -- the closed form
above, evaluated per simulated participant -- against Table 2. This
isolates the model from any numerical-integration or
sampling-schedule effect.

```{r exact-vs-table2}
geomean <- function(x) exp(mean(log(x)))

exact_t2 <- ind |>
  dplyr::group_by(wt_group) |>
  dplyr::summarise(n = dplyr::n(),
                   `Mean WT (kg)` = mean(WT),
                   Selexipag = geomean(auc_p),
                   `JNJ-68006861` = geomean(auc_m),
                   .groups = "drop") |>
  tidyr::pivot_longer(c(Selexipag, `JNJ-68006861`),
                      names_to = "Analyte", values_to = "Exact model") |>
  dplyr::left_join(
    tibble::tribble(
      ~wt_group,        ~Analyte,        ~Published,
      ">=50 kg",        "Selexipag",           9.44,
      ">=25 to <50 kg", "Selexipag",           8.03,
      ">=9 to <25 kg",  "Selexipag",           8.42,
      ">=50 kg",        "JNJ-68006861",       22.20,
      ">=25 to <50 kg", "JNJ-68006861",       19.50,
      ">=9 to <25 kg",  "JNJ-68006861",       17.80
    ), by = c("wt_group", "Analyte")) |>
  dplyr::mutate(`% diff` = 100 * (`Exact model` / Published - 1))

exact_t2 |>
  dplyr::rename("Body weight group" = wt_group) |>
  knitr::kable(digits = 2,
               caption = "Exact closed-form AUCtau,ss (ng*h/mL) by body-weight group vs. Axelsen 2024 Table 2.")

# The metabolite drives the combined metric (37/38 of it) and is the
# endpoint the dose regimen was built on.
#
# Gate design: assert on the MEAN across weight groups, because that is
# what a transcription error moves -- a wrong Vm, km, kmet or reference
# weight shifts all three groups together by tens of percent. The
# per-group scatter is Monte-Carlo noise: with omega(CL) = 0.80 and
# ~100-300 simulated participants per group, the standard error of a
# group's geometric mean is already ~5%, and rxSetSeed() fixes the draw
# only within an rxode2 version, not across the versions CI and a
# developer may resolve. A per-group max with a 15% threshold would
# therefore pass locally and fail CI; the max below is kept only as a
# gross-error backstop, with roughly 2x headroom over the observed noise.
gate_m <- exact_t2$`% diff`[exact_t2$Analyte == "JNJ-68006861"]
cat(sprintf("JNJ-68006861 vs Table 2: mean %+.1f%%, max |diff| %.1f%%\n",
            mean(gate_m), max(abs(gate_m))))
stopifnot(abs(mean(gate_m)) < 15, max(abs(gate_m)) < 30)
```

## Steady-state concentration-time profiles by body-weight group

Replicates Figure S1 of Axelsen 2024 (mean dose-normalised plasma
concentration-time profiles of selexipag and its active metabolite by
body-weight group), on a dense grid.

```{r figure-s1, fig.width = 8, fig.height = 4}
rxode2::rxSetSeed(20241213)
set.seed(20241213)

dense <- rxode2::rxSolve(
  mod_ped,
  make_events(cohort, seq(0, tau, by = 0.25)),
  keep = keep_cols,
  returnType = "data.frame"
) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(time = time - (n_doses - 1) * tau)

dense |>
  dplyr::select(time, wt_group, dose_ug, Cc, Cc_act) |>
  tidyr::pivot_longer(c(Cc, Cc_act), names_to = "analyte", values_to = "conc") |>
  dplyr::mutate(
    analyte = factor(analyte, c("Cc", "Cc_act"),
                     c("Selexipag", "JNJ-68006861")),
    conc_dn = conc / dose_ug * dplyr::case_when(
      wt_group == ">=9 to <25 kg" ~ 100,
      wt_group == ">=25 to <50 kg" ~ 150,
      TRUE ~ 200)
  ) |>
  dplyr::group_by(analyte, wt_group, time) |>
  dplyr::summarise(mean = mean(conc_dn), .groups = "drop") |>
  ggplot2::ggplot(ggplot2::aes(time, mean, colour = wt_group)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::facet_wrap(~analyte, scales = "free_y") +
  ggplot2::labs(x = "Time after dose (h)",
                y = "Mean dose-normalised concentration (ng/mL)",
                colour = "Body weight group",
                title = "Replicates Figure S1 of Axelsen 2024") +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

# PKNCA validation

```{r pknca}
nca_input <- sim |>
  dplyr::select(id, time, wt_group, dose_ug, Cc, Cc_act) |>
  tidyr::pivot_longer(c(Cc, Cc_act), names_to = "analyte", values_to = "conc") |>
  dplyr::mutate(analyte = dplyr::recode(analyte,
                                        Cc = "Selexipag",
                                        Cc_act = "JNJ-68006861")) |>
  dplyr::filter(!is.na(conc))

dose_input <- cohort |>
  dplyr::select(id, wt_group, dose_ug) |>
  tidyr::crossing(analyte = c("Selexipag", "JNJ-68006861")) |>
  dplyr::mutate(time = 0)

conc_obj <- PKNCA::PKNCAconc(
  nca_input,
  conc ~ time | analyte + wt_group + id,
  concu = "ng/mL", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  dose_input,
  dose_ug ~ time | analyte + wt_group + id,
  doseu = "ug"
)

intervals <- data.frame(start = 0, end = tau,
                        cmax = TRUE, tmax = TRUE, auclast = TRUE)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)

nca_tbl <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast"))

# Each participant is simulated at their own starting dose, so the raw
# NCA output is already on the dose-normalisation basis Table 2 uses
# (100 / 150 / 200 ug for the three weight bands); no rescaling applies.
# tmax is not dose-normalised in Table 2 either (Table 2 note a).
sim_summary <- nca_tbl |>
  dplyr::group_by(analyte, wt_group, PPTESTCD) |>
  dplyr::summarise(
    PPORRES = dplyr::if_else(dplyr::first(PPTESTCD) == "tmax",
                             stats::median(PPORRES), geomean(PPORRES)),
    .groups = "drop"
  )
```

## Comparison against the published NCA (Table 2)

Table 2 of Axelsen 2024 reports dose-normalised steady-state NCA by
body-weight group: geometric-mean `Cmax,ss,dn` and `AUCtau,ss,dn`, and
median `tmax,ss` (not dose-normalised).

```{r table2-comparison}
reference_t2 <- tibble::tribble(
  ~analyte,        ~wt_group,        ~cmax, ~tmax, ~auclast,
  "Selexipag",     ">=50 kg",         4.06,  1.00,     9.44,
  "Selexipag",     ">=25 to <50 kg",  2.97,  1.05,     8.03,
  "Selexipag",     ">=9 to <25 kg",   2.75,  2.00,     8.42,
  "JNJ-68006861",  ">=50 kg",         3.87,  3.01,    22.20,
  "JNJ-68006861",  ">=25 to <50 kg",  3.32,  2.02,    19.50,
  "JNJ-68006861",  ">=9 to <25 kg",   3.05,  3.04,    17.80
)

cmp_t2 <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_summary,
  reference = reference_t2,
  by = c("analyte", "wt_group"),
  params = c("cmax", "tmax", "auclast"),
  units = c(cmax = "ng/mL", tmax = "h", auclast = "ng*h/mL")
)

knitr::kable(
  dplyr::rename(cmp_t2, "Analyte" = analyte, "Body weight group" = wt_group),
  digits = 2,
  caption = "Simulated vs. Axelsen 2024 Table 2 (dose-normalised steady-state NCA by body-weight group)."
)
attr(cmp_t2, "footnote")
```

```{r table2-gate}
chk_t2 <- cmp_t2 |>
  dplyr::mutate(pct = suppressWarnings(as.numeric(gsub("[^0-9.+-]", "", `% diff`)))) |>
  dplyr::filter(!is.na(pct))

# tmax is excluded from the numeric gate. Both the published and the
# simulated tmax are medians over a 7-point grid, so each can only take
# grid values; the published medians additionally carry real sampling-time
# jitter (the paper reports ranges like 0.92-8.00 h) that a simulation with
# nominal times cannot reproduce. A "% diff" between two grid-quantised
# medians is an artefact, not a model check.
chk_t2_exposure <- dplyr::filter(chk_t2, !grepl("^Tmax", `NCA parameter`))

# Cohort-level comparison: assert on the CENTRE and a robust quantile, not
# on the extremes. A mis-transcribed clearance, volume, dose or unit moves
# the whole distribution by tens of percent and breaks the median gate; the
# scatter that survives is cohort-composition and Monte-Carlo noise on top
# of the paper's own N = 12-28 per group.
cat(sprintf("Cmax/AUC vs Table 2: median %+.1f%%, 90th pct |diff| %.1f%%, max |diff| %.1f%%\n",
            stats::median(chk_t2_exposure$pct),
            stats::quantile(abs(chk_t2_exposure$pct), 0.9),
            max(abs(chk_t2_exposure$pct))))
stopifnot(
  abs(stats::median(chk_t2_exposure$pct)) < 20,
  stats::quantile(abs(chk_t2_exposure$pct), 0.9) < 50
)
```

Reading this table:

* **JNJ-68006861** -- the analyte that carries 37/38 of the combined
  potency-weighted exposure the dose regimen was built on -- tracks the
  published geometric means within about 10% in every weight group, for
  both `Cmax` and `AUCtau,ss`.
* **Selexipag** matches closely in the two lighter groups (`AUClast`
  -0.1% and -3.3%) but the model under-predicts the heaviest group
  (`>=50 kg`, -33%). Two things contribute. First, the published Table 2
  selexipag `AUCtau,ss,dn` is **not monotone in body weight** (8.42,
  8.03, 9.44 ng*h/mL for increasing weight) whereas the model's weight
  effect is strictly monotone decreasing, so no parameterisation of this
  model can match all three; the exact-model table above shows the same
  pattern (9.13, 8.24, 7.15) without any NCA involved. Second, that
  group is the paper's smallest (N = 12) and its published 95% CI is
  wide (7.32-12.2 ng*h/mL). Second-order, the sparse 7-point grid costs
  a further ~10% of the true AUC, and it costs the same ~10% in the
  published NCA, so it largely cancels.
* `tmax` rows are shown for completeness but are excluded from the gate,
  for the reason given in the chunk above.

None of this affects the paper's conclusion, which rests on the
combined potency-weighted exposure: that metric is reproduced to within
0.3% overall in the next section.

# Model-based combined exposure (Table 3)

Table 3 reports the geometric mean of the model-based
`AUCtau,ss,combined` in adults (19.0, N = 510) and in pediatric
participants overall and by starting-dose group and age cohort, plus
the pediatric/adult geometric-mean ratio. Here the same quantity is
computed per simulated participant from their individual parameters,
which is Axelsen 2024's own procedure.

```{r table3-comparison}
adult_ref_gm <- 19.0   # Axelsen 2024 Table 3, adults, N = 510

t3_sim <- dplyr::bind_rows(
  ind |> dplyr::summarise(Stratifier = "Population", Group = "Pediatrics",
                          gm = geomean(auc_comb)),
  ind |> dplyr::group_by(Group = dose_group) |>
    dplyr::summarise(gm = geomean(auc_comb), .groups = "drop") |>
    dplyr::mutate(Stratifier = "Pediatric starting dose group"),
  ind |> dplyr::group_by(Group = cohort) |>
    dplyr::summarise(gm = geomean(auc_comb), .groups = "drop") |>
    dplyr::mutate(Stratifier = "Cohort")
) |>
  dplyr::mutate(GMR_sim = gm / adult_ref_gm)

t3_pub <- tibble::tribble(
  ~Group,        ~gm_pub, ~gmr_pub,
  "Pediatrics",     19.6,    1.03,
  "100 ug",         18.4,   0.966,
  "150 ug",         19.9,    1.05,
  "200 ug",         22.3,    1.17,
  "2-5 years",      18.0,   0.949,
  "6-11 years",     19.8,    1.04,
  "12-17 years",    20.8,    1.09
)

t3 <- t3_sim |>
  dplyr::left_join(t3_pub, by = "Group") |>
  dplyr::mutate(pct_diff = 100 * (gm / gm_pub - 1)) |>
  dplyr::select(Stratifier, Group, gm_pub, gm, gmr_pub, GMR_sim, pct_diff)

t3 |>
  dplyr::rename("Published geo-mean" = gm_pub, "Simulated geo-mean" = gm,
                "Published GMR vs adults" = gmr_pub,
                "Simulated GMR vs adults" = GMR_sim, "% diff" = pct_diff) |>
  knitr::kable(digits = 3,
               caption = "Model-based AUCtau,ss,combined (ng*h/mL) vs. Axelsen 2024 Table 3.")
```

```{r table3-gate}
stopifnot(
  abs(stats::median(t3$pct_diff)) < 15,
  stats::quantile(abs(t3$pct_diff), 0.9) < 30,
  # The paper's headline conclusion: pediatric exposure matches adult
  # exposure once body weight is accounted for. The published GMRs span
  # 0.949-1.17; the simulated ones must land in the same neighbourhood.
  all(t3$GMR_sim > 0.7 & t3$GMR_sim < 1.5)
)
```

## Typical-value reproduction of the adult reference

A single typical 70 kg adult, evaluated across the four
PAH-comedication strata, brackets the published adult geometric mean of
19.0 -- and the modal stratum (ERA + PDE5, which covers roughly half
the pediatric cohort and the majority of GRIPHON) lands within a few
percent of it. This is the cohort-composition-free version of the check
above.

```{r adult-typical}
adult_strata <- tibble::tibble(
  id = 1:4, dose_ug = 200, WT = 70, TBILI = 10, SEXF = 1,
  CONMED_ERA       = c(0, 1, 0, 0),
  CONMED_PDE5I     = c(0, 0, 1, 0),
  CONMED_ERA_PDE5I = c(0, 0, 0, 1),
  stratum = c("PAH-naive", "ERA only", "PDE5I only", "ERA + PDE5")
)

adult_typ <- rxode2::rxSolve(
  rxode2::zeroRe(mod_adult),
  make_events(adult_strata, 0),
  keep = c("stratum", "dose_ug"),
  returnType = "data.frame"
) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(auc_comb = auc_combined(dose_ug, vc, kmet, cl, vc_act, km_act),
                pct_vs_pub = 100 * (auc_comb / adult_ref_gm - 1))

adult_typ |>
  dplyr::select(stratum, auc_comb, pct_vs_pub) |>
  dplyr::rename("PAH comedication stratum" = stratum,
                "AUCtau,ss,combined" = auc_comb,
                "% vs published adult geo-mean (19.0)" = pct_vs_pub) |>
  knitr::kable(digits = 2,
               caption = "Typical 70 kg adult female on 200 ug b.i.d., adult model.")

# The published adult geometric mean must sit inside the range spanned by
# the four comedication strata: a transcription error in CL, Vp, kmet, km,
# Vm or the potency weights would move the whole set off 19.0.
stopifnot(
  min(adult_typ$auc_comb) < adult_ref_gm,
  max(adult_typ$auc_comb) > adult_ref_gm,
  # ERA + PDE5 is the modal stratum and should be the closest match.
  abs(adult_typ$pct_vs_pub[adult_typ$stratum == "ERA + PDE5"]) < 10
)
```

# Adult vs. pediatric parameter comparison

Reproduces the parameter-consistency claims made in Axelsen 2024
Results ("Population pharmacokinetics analysis ... in pediatric
participants").

```{r param-comparison, echo = FALSE}
tibble::tribble(
  ~Parameter, ~Adult, ~Pediatric, ~Comment,
  "beta_CL(WEIGHTBL)", 0.546, 0.828, "Paper: 'close to the standard value of 0.75'",
  "beta_Vp(WEIGHTBL)", 1.04,  0.805, "Paper: 'close to the previous adult value, as well as the standard value (1.0)'",
  "beta_Vm(WEIGHTBL)", 0.803, 0.542, "Paper: '~30% smaller than the previous adult value'",
  "CL (L/h)",          18,    18.9,  "Consistent",
  "Vp (L)",            12.2,  12.4,  "Consistent",
  "Vm (L)",            5.88,  6.05,  "Consistent"
) |>
  knitr::kable(caption = "Adult vs. pediatric estimates, Axelsen 2024 Table 1.")
```

# Assumptions, deviations and errata

**Assumptions made in this vignette (not in the paper).**

* The virtual cohort's continuous covariates are drawn log-normal
  matched to the Table S3 mean and SD and truncated to the published
  range. The paper does not report the distributional shape; log-normal
  is used because both weight and bilirubin are strictly positive and
  right-skewed (bilirubin markedly so: mean 15.1, SD 16.8, max 84 in
  the oldest cohort).
* Sex and PAH-comedication stratum are drawn independently of weight
  and of each other. The paper reports only marginal proportions, so
  any correlation structure is unknown.
* Starting dose is derived from the protocol's body-weight bands rather
  than drawn from the observed dose distribution.
* 200 participants are simulated per age cohort rather than the
  published 17-21, to give a stable estimate of the model's predicted
  geometric mean. Cohort-level agreement is therefore assessed on
  medians and robust quantiles, never on extremes.
* The adult reference geometric mean (19.0) is taken as the published
  constant from Table 3 rather than re-simulated, because Axelsen 2024
  does not tabulate the adult covariate distribution.

**Errata / apparent typographical slips in the source.**

* Table 1 gives the units of `CL` as "(1/h)". `CL` is a clearance and
  must be in L/h: the printed exposure formula adds it to `Vp * kmet`
  (L x 1/h = L/h), and `Dose(ug) / CL(L/h)` gives ng/mL only if `CL` is
  in L/h. Encoded as L/h.
* Table 1 gives the units of `error_PROP2` as "(ug/mL)". Both analytes
  are measured in ng/mL: the control stream header declares
  "Concentration: ng/mL" globally, and Table 2 reports metabolite
  `Cmax,ss,dn` in ng/mL. In any case the parameter is a *proportional*
  (unitless fraction) error, so the unit annotation is doubly spurious.
* Table 1 lists `Tlag1Half` with units "(h)". The quantity that is
  logit-transformed and carries the LogitNormal IIV is dimensionless
  (bounded on 0-1); the lag time in hours is `2 * Tlag1Half`. The
  packaged model carries the parameter on the logit scale as
  `logittlag` and forms `tlag = 2 * expit(logittlag + eta)`, which
  reproduces both the typical value (0.668 h) and the (0, 2) h bound.

**Modelling notes.**

* The metabolite-formation flux transfers *mass* one-for-one
  (`k2T4 = kmet` in the control stream), with no molecular-weight
  correction between selexipag (MW 496.6) and JNJ-68006861 (MW 442.5).
  Any such factor is absorbed into the apparent metabolite volume
  `Vm/F`, which is what the model estimates. Metabolite concentrations
  are therefore on the paper's own apparent scale and are directly
  comparable to its reported values.
* The reference body weight is 70 kg, well above every pediatric
  participant, so the weight terms extrapolate downwards throughout the
  pediatric model. This is the authors' choice, retained here so the
  coefficients are used exactly as published.
* `omega(kmet)` and `omega(Vm)` have 76% and 80% shrinkage in the
  pediatric fit, so those IIV terms are weakly informed by the
  pediatric data and effectively carried from the adult prior.
* `units$dosing` is `ug` while `units$concentration` is `ng/mL`. This
  is self-consistent and needs no scaling: volumes are in L, so
  `ug / L = ng/mL` exactly. The convention checker raises an
  informational note about the magnitude difference; it is expected.
* The sibling model `Krause_2017_selexipag` (the upstream Monolix fit
  of the same adult data) encodes its categorical covariate effects in
  the `(1 + beta * I)` form, whereas this extraction uses
  `exp(beta * I)` because Axelsen's control stream adds each
  coefficient to the MU-referenced log-scale parameter before
  exponentiating. The coefficients are numerically close, so the two
  forms agree to within 1-2%.
