---
title: "Telisotuzumab vedotin conjugate, MMAE payload and exposure-response (Babel 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Telisotuzumab vedotin conjugate, MMAE payload and exposure-response (Babel 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Source and models

```{r models, include = FALSE}
mod_adc  <- readModelDb("Babel_2026_telisotuzumab")
mod_mmae <- readModelDb("Babel_2026_telisotuzumab_mmae")
mod_orr  <- readModelDb("Babel_2026_telisotuzumab_orr")
mod_pn   <- readModelDb("Babel_2026_telisotuzumab_neuropathy")
mod_ce   <- readModelDb("Babel_2026_telisotuzumab_corneal")
mod_teae <- readModelDb("Babel_2026_telisotuzumab_teae")
meta_adc <- rxode2::rxode2(mod_adc)$meta
```

* Citation: `r meta_adc$reference`
* Article (open access): <https://doi.org/10.1002/psp4.70219>
* PubMed Central: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC12945708/>
* Supporting Information (Data S1; Tables S1-S7 and Figures S1-S5): distributed
  with the article and retrieved from the Europe PMC supplementary-files
  endpoint for PMC12945708.

Babel 2026 reports **six** models, and this package carries all six as separate
files because the authors fitted them separately:

| Model file | What it is | Source |
|---|---|---|
| `Babel_2026_telisotuzumab` | Conjugate (Teliso-V) population PK, 2-compartment | Table S7 conjugate block, Figure S1A |
| `Babel_2026_telisotuzumab_mmae` | Unconjugated MMAE payload population PK, 1-compartment with first-order deconjugation | Table S7 payload block, Figure S1B |
| `Babel_2026_telisotuzumab_orr` | Exposure-efficacy logistic for ORR per Independent Central Review | Figure 2 |
| `Babel_2026_telisotuzumab_neuropathy` | Exposure-safety logistic for grade >= 3 peripheral neuropathy | Figure 4 left |
| `Babel_2026_telisotuzumab_corneal` | Exposure-safety logistic for grade >= 2 corneal epitheliopathy | Figure 4 right |
| `Babel_2026_telisotuzumab_teae` | Exposure-safety logistic for any grade >= 3 TEAE, driven by **payload** | Figure 5, CavgMMAE panel |

The conjugate and payload PK models were "developed independently since
independent models show similar performance compared to integrated models"
(Methods). Figure S1 draws them as two separate panels, and the payload panel
shows a single MMAE compartment with an inbound first-order arrow labelled
`Ka` and an outbound clearance arrow -- not a link from the conjugate
compartments. That is why `Babel_2026_telisotuzumab_mmae` is a standalone
depot-plus-central model rather than a coupled system, and why it can be solved
without the conjugate model.

# Population

```{r population, echo = FALSE}
pop <- meta_adc$population
knitr::kable(
  tibble::tibble(
    Field = c("Species", "Subjects", "Studies", "Age", "Body weight",
              "Female", "Race", "Disease", "Doses", "Regions"),
    Value = c(pop$species, as.character(pop$n_subjects), as.character(pop$n_studies),
              pop$age_range, pop$weight_range, paste0(pop$sex_female_pct, "%"),
              paste(names(pop$race_ethnicity), pop$race_ethnicity, sep = " ",
                    collapse = "%, "),
              pop$disease_state, pop$dose_range, pop$regions)
  ),
  caption = "Population of the Babel 2026 population PK analysis (Table S4)."
)
```

The population PK analysis pooled 304 patients: 35 from the phase 1 study
NCT02099058 (monotherapy cohorts that received Process II drug material, all
solid tumour types) and 269 from the LUMINOSITY phase 2 study NCT03539536
(c-Met protein overexpressing NSCLC). The exposure-efficacy analysis is a
193-patient subset of LUMINOSITY only -- the phase 1 study was excluded because
its c-Met overexpression criteria differed -- and the exposure-safety analysis
uses 284 patients pooled across both studies.

The three covariate medians that all six PK covariate terms are normalised to
come from Table S4 (All Participants, N = 304) and are independently confirmed
by Table S3, whose forest-plot reference strata are split at exactly those
values: body weight 68.9 kg, baseline albumin 41.6 g/L and age 65 years.

# Source trace

Every value in the six `ini()` blocks, with where it came from.

```{r source-trace, echo = FALSE}
knitr::kable(
  tibble::tribble(
    ~Model,        ~Parameter,           ~Value,        ~`Source location`,
    "conjugate",   "lcl",                "1.34 L/day",  "Table S7, 'CL (L/day)'",
    "conjugate",   "lvc",                "3.44 L",      "Table S7, 'Vc (L)'",
    "conjugate",   "lq",                 "1.21 L/day",  "Table S7, 'Q (L/day)'",
    "conjugate",   "lvp",                "2.41 L",      "Table S7, 'Vp (L)'",
    "conjugate",   "e_wt_cl_q",          "0.405",       "Table S7, 'Body Weight on CL and Q'",
    "conjugate",   "e_wt_vc_vp",         "0.590",       "Table S7, 'Body Weight on Vc and Vp'",
    "conjugate",   "e_ada_cl",           "1.17",        "Table S7, 'ADA on CL'",
    "conjugate",   "e_alb_cl",           "-0.970",      "Table S7, 'Albumin on CL'",
    "conjugate",   "e_black_cl",         "0.763",       "Table S7, 'Black or African American vs. White on CL'",
    "conjugate",   "e_asian_cl",         "0.887",       "Table S7, 'Asian vs. White on CL'",
    "conjugate",   "e_age_vc",           "0.223",       "Table S7, 'Age on Vc'",
    "conjugate",   "e_alb_vc",           "-0.464",      "Table S7, 'Albumin on Vc'",
    "conjugate",   "e_sexf_vc",          "0.925",       "Table S7, 'Sex on Vc'",
    "conjugate",   "etalcl, etalvc",     "0.0946, 0.0268", "Table S7, 'IIV on CL' / 'IIV on Vc' (variances)",
    "conjugate",   "propSd, addSd",      "sqrt(0.0528), sqrt(0.0447)", "Table S7, error rows (variances)",
    "payload",     "lka",                "0.154 /day",  "Table S7, 'Ka (1/day)'",
    "payload",     "lcl",                "76.3 L/day",  "Table S7, 'CL (L/day)'",
    "payload",     "lvc",                "96.0 L",      "Table S7, 'Vc (L)'",
    "payload",     "e_alb_cl",           "1.99",        "Table S7, 'Albumin on CL'",
    "payload",     "e_renalmild_cl",     "0.842",       "Table S7, 'Mild vs. Normal Renal Impairment on CL'",
    "payload",     "e_renalmodsev_cl",   "0.755",       "Table S7, 'Moderate/Severe vs. Normal Renal Impairment on CL'",
    "payload",     "e_age_ka",           "-0.454",      "Table S7, 'Age on Ka'",
    "payload",     "e_alb_ka",           "-1.01",       "Table S7, 'Albumin on Ka'",
    "payload",     "e_black_ka",         "0.853",       "Table S7, 'Black or African American vs. White on Ka'",
    "payload",     "e_asian_ka",         "0.874",       "Table S7, 'Asian vs. White on Ka'",
    "payload",     "e_wt_vc",            "0.612",       "Table S7, 'Body Weight on Vc'",
    "payload",     "etalcl, etalvc, etalka", "0.240, 0.486, 0.0735", "Table S7, IIV rows (variances)",
    "payload",     "propSd, addSd",      "sqrt(0.0833), sqrt(7.53e-10)", "Table S7, error rows (variances)",
    "ORR",         "logit_ref, e_cav_orr",  "-5.292, 2.450",  "DIGITISED, Figure 2 fitted line",
    "neuropathy",  "logit_ref, e_cav_pn",   "-7.437, 2.793",  "DIGITISED, Figure 4 left fitted line",
    "corneal",     "logit_ref, e_cav_ce",   "-9.786, 4.206",  "DIGITISED, Figure 4 right fitted line",
    "TEAE",        "logit_ref, e_cav_teae", "-1.0835, 0.7195", "DIGITISED, Figure 5 CavgMMAE fitted line",
    "all ER",      "addSd_prob_*",       "fixed(0)-like 0.001", "NOT from source; placeholder so rxode2 accepts an observation"
  ),
  caption = "Source trace for every ini() value in the six model files."
)
```

Every PK model equation in `model()` is either the standard two- or
one-compartment mass-balance system drawn in Figure S1, or a covariate term of
the form the Methods prescribes: "Continuous covariates were normalized to the
median of the overall population and incorporated into the model using a power
function. Categorical covariates were included multiplicatively to obtain the
proportional change between the tested categorical groups." The only hardcoded
constants are the three normalisation medians (68.9 kg, 41.6 g/L, 65 years),
all sourced above.

# Check 1 -- the IIV column of Table S7 is a variance, not a standard deviation

Table S7 prints a "Population Estimate" and a "%CV" for each random effect, and
its footnote states the relationship: "%CV was calculated as
`SQRT(exp(omega2)-1)*100`". That footnote makes the table **self-pinning**: if
the Population Estimate column is the variance `omega^2`, the formula must
reproduce the printed %CV exactly. It does, for all five random effects, which
is why the `ini()` eta lines carry the printed numbers unchanged rather than
squared.

```{r check-iiv}
iiv <- tibble::tribble(
  ~Model,      ~Parameter, ~omega2,  ~cv_printed,
  "conjugate", "CL",       0.0946,   31.5,
  "conjugate", "Vc",       0.0268,   16.5,
  "payload",   "CL",       0.240,    52.1,
  "payload",   "Vc",       0.486,    79.1,
  "payload",   "Ka",       0.0735,   27.6
) |>
  dplyr::mutate(
    cv_from_variance = sqrt(exp(omega2) - 1) * 100,
    cv_if_it_were_sd = sqrt(exp(omega2^2) - 1) * 100
  )

knitr::kable(iiv, digits = 3,
             caption = "Table S7 footnote formula applied both ways.")

stopifnot(
  # The variance reading reproduces every printed %CV to the printed precision.
  all(abs(iiv$cv_from_variance - iiv$cv_printed) < 0.05),
  # The standard-deviation reading does not, for any of them.
  all(abs(iiv$cv_if_it_were_sd - iiv$cv_printed) > 1)
)
```

# Check 2 -- Table S7 point estimates are consistent with their own %RSE and 95% CI

A transcription check on the structural parameters: for a Wald interval the
95% CI half-width should be about 1.96 standard errors, and the standard error
is `%RSE / 100` times the estimate. Agreement confirms that the estimate, the
%RSE and the interval were all read from the same row.

```{r check-rse}
rse <- tibble::tribble(
  ~Model,      ~Parameter, ~est,   ~rse_pct, ~lo,    ~hi,
  "conjugate", "CL",       1.34,   2.98,     1.26,   1.42,
  "conjugate", "Vc",       3.44,   1.38,     3.35,   3.54,
  "conjugate", "Q",        1.21,   4.59,     1.11,   1.33,
  "conjugate", "Vp",       2.41,   2.68,     2.28,   2.54,
  "payload",   "CL",       76.3,   4.83,     69.4,   83.9,
  "payload",   "Vc",       96.0,   5.62,     86.0,   107,
  "payload",   "Ka",       0.154,  2.58,     0.147,  0.162
) |>
  dplyr::mutate(
    halfwidth_ci   = (hi - lo) / 2,
    halfwidth_wald = 1.96 * rse_pct / 100 * est,
    ratio          = halfwidth_ci / halfwidth_wald
  )

knitr::kable(rse, digits = 4,
             caption = "Printed 95% CI half-width against 1.96 x SE from the printed %RSE.")

stopifnot(
  # Centre: a mis-transcribed estimate or %RSE would move the whole column.
  abs(median(rse$ratio) - 1) < 0.10,
  # Envelope: robust to one asymmetric or profile-likelihood interval.
  quantile(abs(rse$ratio - 1), 0.9) < 0.25
)
```

# Virtual cohort

The cohort reproduces the marginal covariate distributions of Table S4 (All
Participants). Continuous covariates are drawn from truncated normals with the
printed mean, standard deviation and min/max; categorical covariates from the
printed proportions. Two arms of 200 subjects each, at the two LUMINOSITY dose
levels, with the same covariate draws in both arms so that the dose comparison
is paired.

```{r cohort}
set.seed(20260212)
n_sub <- 200L

rtrunc_norm <- function(n, mean, sd, lo, hi) {
  x <- stats::rnorm(n, mean, sd)
  pmin(pmax(x, lo), hi)
}

race <- sample(c("White", "Asian", "Black"), n_sub, replace = TRUE,
               prob = c(0.65, 0.32, 0.03))
renal <- sample(c("normal", "mild", "moderate", "severe"), n_sub, replace = TRUE,
                prob = c(0.36, 0.43, 0.19, 0.01))

cohort <- tibble::tibble(
  id            = seq_len(n_sub),
  WT            = rtrunc_norm(n_sub, 70.7, 16.7, 36.0, 144),   # Table S4 mean (SD), min/max
  AGE           = rtrunc_norm(n_sub, 63,   10,   30,   87),    # Table S4
  ALB           = rtrunc_norm(n_sub, 40.9, 4.40, 29.0, 52.0),  # Table S4
  SEXF          = stats::rbinom(n_sub, 1, 0.378),              # Table S4, 115/304 female
  RACE_BLACK    = as.integer(race == "Black"),
  RACE_ASIAN    = as.integer(race == "Asian"),
  ADA_POS       = stats::rbinom(n_sub, 1, 61 / 304),           # Figure 1A, 61 positive of 304
  RENALIMP_MILD = as.integer(renal == "mild"),
  RENALIMP_MOD  = as.integer(renal == "moderate"),
  RENALIMP_SEV  = as.integer(renal == "severe")
)

# 1.9 mg/kg, capped at 190 mg for patients of at least 100 kg (Abstract).
teliso_dose <- function(wt, mgkg) pmin(mgkg * wt, 190)

knitr::kable(
  cohort |>
    dplyr::summarise(
      `WT median`  = median(WT), `AGE median` = median(AGE),
      `ALB median` = median(ALB), `Female %`  = 100 * mean(SEXF),
      `Asian %`    = 100 * mean(RACE_ASIAN), `ADA+ %` = 100 * mean(ADA_POS)
    ),
  digits = 1, caption = "Simulated cohort against the Table S4 medians 68.9 kg, 65 years, 41.6 g/L."
)
```

# Conjugate concentration-time profiles

Eight every-2-week doses, an assumed 30-minute infusion (Babel 2026 states
"intravenous infusion" and its PK sampling schedule references "the end of
infusion" but never prints a duration -- see Assumptions below; the conjugate
has no absorption step, so the duration affects only the very first minutes of
the peak).

```{r sim-adc}
tau      <- 14      # days
n_dose   <- 8L
inf_dur  <- 0.5 / 24  # 30 min, expressed in days

sim_arm <- function(mgkg, seed) {
  rxode2::rxSetSeed(seed)   # common random numbers per arm
  amt <- teliso_dose(cohort$WT, mgkg)
  dose <- data.frame(
    id   = rep(cohort$id, each = n_dose),
    time = rep(seq(0, by = tau, length.out = n_dose), times = n_sub),
    amt  = rep(amt, each = n_dose),
    rate = rep(amt / inf_dur, each = n_dose),
    evid = 1L,
    cmt  = "central"
  )
  # Dense observations over cycle 1 and over the eighth (near-steady-state)
  # interval; cmt is the ODE STATE, and rxode2 returns Cc alongside it.
  # The grid is deliberately fine over the first day of each interval: the
  # 30-minute infusion and the fast distribution phase make a 0.25-day
  # trapezoid underestimate the early AUC by several percent, which would
  # show up as a false failure of the mass-balance identity below.
  fine <- c(seq(0, 0.06, by = 0.005), seq(0.08, 1, by = 0.02))
  obs_t <- c(fine, seq(1.25, tau, by = 0.25),
             (n_dose - 1) * tau + fine,
             seq((n_dose - 1) * tau + 1.25, n_dose * tau, by = 0.25))
  obs <- data.frame(
    id   = rep(cohort$id, each = length(obs_t)),
    time = rep(obs_t, times = n_sub),
    amt  = 0, rate = 0, evid = 0L, cmt = "central"
  )
  ev <- dplyr::bind_rows(dose, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    dplyr::left_join(cohort, by = "id")
  out <- as.data.frame(rxode2::rxSolve(mod_adc, events = ev, returnType = "data.frame"))
  out$dose_mgkg <- mgkg
  out
}

sim19 <- sim_arm(1.9, 20260212L)
sim16 <- sim_arm(1.6, 20260212L)
sim_adc <- dplyr::bind_rows(sim19, sim16)
```

```{r plot-adc, echo = FALSE, fig.cap = "Simulated Teliso-V conjugate concentrations, median and 5th-95th percentile band, both LUMINOSITY dose levels. Compare with the prediction-corrected VPC of Figure S4."}
sim_adc |>
  dplyr::filter(time <= tau) |>
  dplyr::group_by(dose_mgkg, time) |>
  dplyr::summarise(
    lo = quantile(Cc, 0.05), md = median(Cc), hi = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot2::ggplot(ggplot2::aes(time, md, colour = factor(dose_mgkg),
                               fill = factor(dose_mgkg))) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi), alpha = 0.18, colour = NA) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::scale_y_log10() +
  ggplot2::labs(x = "Time after first dose (day)", y = "Conjugate concentration (ug/mL)",
                colour = "Dose (mg/kg)", fill = "Dose (mg/kg)") +
  ggplot2::theme_bw()
```

# Check 3 -- mass balance: AUC(0, inf) x CL equals the dose exactly

A closed-form identity that holds for any linear compartmental system at any
time: the amount eliminated up to time T equals the dose in minus the amount
still in the body. Evaluated per subject on the same solve that produced the
profiles, so the two sides are numerically independent only through the ODE
solver -- which is exactly what this checks. A tight bound is correct here
because the only source of disagreement is integration error.

```{r check-massbalance}
mb <- sim19 |>
  dplyr::filter(time <= tau) |>
  dplyr::group_by(id) |>
  dplyr::summarise(
    auc     = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    remain  = dplyr::last(central) + dplyr::last(peripheral1),
    cl_i    = dplyr::first(cl),        # the individual's own clearance
    .groups = "drop"
  ) |>
  dplyr::left_join(cohort, by = "id") |>
  dplyr::mutate(
    dose    = teliso_dose(WT, 1.9),
    balance = (auc * cl_i + remain) / dose
  )

stopifnot(
  abs(median(mb$balance) - 1) < 0.01,
  quantile(abs(mb$balance - 1), 0.9) < 0.02
)
knitr::kable(
  tibble::tibble(
    Quantity = c("median (eliminated + remaining) / dose", "90th pct |deviation|"),
    Value    = c(median(mb$balance), quantile(abs(mb$balance - 1), 0.9))
  ),
  digits = 4,
  caption = "Mass balance of the conjugate model over cycle 1, per subject."
)
```

This identity uses each subject's own realised clearance, so it tests the ODE
system and the observation equation `Cc = central / vc` but not the covariate
algebra. The next check does that separately, and deterministically.

# Check 3b -- covariate algebra reproduces the Table S7 formulas exactly

Solving with the random effects zeroed makes each subject's `cl`, `vc`, `q` and
`vp` deterministic functions of the covariates, so they can be compared against
the Table S7 numbers recomputed independently here. A tight bound is correct:
the two sides should agree to machine precision or the model file has a wrong
exponent, a wrong reference value, or a covariate attached to the wrong
parameter.

```{r check-covariate-algebra}
ev_typ <- dplyr::bind_rows(
  data.frame(id = cohort$id, time = 0, amt = 1, rate = 1 / inf_dur,
             evid = 1L, cmt = "central"),
  data.frame(id = cohort$id, time = 1, amt = 0, rate = 0, evid = 0L, cmt = "central")
) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  dplyr::left_join(cohort, by = "id")

typ <- as.data.frame(
  rxode2::rxSolve(rxode2::zeroRe(mod_adc), events = ev_typ, returnType = "data.frame")
) |>
  dplyr::distinct(WT, AGE, ALB, SEXF, RACE_BLACK, RACE_ASIAN, ADA_POS,
                  cl, vc, q, vp) |>
  dplyr::mutate(
    cl_ref = 1.34 * (WT / 68.9)^0.405 * (ALB / 41.6)^-0.970 *
             1.17^ADA_POS * 0.763^RACE_BLACK * 0.887^RACE_ASIAN,
    vc_ref = 3.44 * (WT / 68.9)^0.590 * (AGE / 65)^0.223 *
             (ALB / 41.6)^-0.464 * 0.925^SEXF,
    q_ref  = 1.21 * (WT / 68.9)^0.405,
    vp_ref = 2.41 * (WT / 68.9)^0.590
  )

stopifnot(
  max(abs(typ$cl / typ$cl_ref - 1)) < 1e-8,
  max(abs(typ$vc / typ$vc_ref - 1)) < 1e-8,
  max(abs(typ$q  / typ$q_ref  - 1)) < 1e-8,
  max(abs(typ$vp / typ$vp_ref - 1)) < 1e-8
)

knitr::kable(
  tibble::tibble(
    Parameter = c("CL", "Vc", "Q", "Vp"),
    `Max relative deviation` = c(
      max(abs(typ$cl / typ$cl_ref - 1)), max(abs(typ$vc / typ$vc_ref - 1)),
      max(abs(typ$q  / typ$q_ref  - 1)), max(abs(typ$vp / typ$vp_ref - 1))
    ),
    `Typical-covariate value` = c(1.34, 3.44, 1.21, 2.41),
    `Model at the medians` = c(
      1.34 * 1, 3.44 * 1, 1.21 * 1, 2.41 * 1
    )
  ),
  digits = 12,
  caption = "Model covariate algebra against the Table S7 formulas, over all 200 simulated covariate vectors."
)
```

# PKNCA analysis of the simulated conjugate profiles

```{r pknca}
conc_c1 <- sim19 |>
  dplyr::filter(time <= tau, !is.na(Cc)) |>
  dplyr::select(id, time, Cc) |>
  dplyr::distinct(id, time, .keep_all = TRUE)

# Defensive time-zero record so PKNCA never asks for an AUC starting before the
# first measurement.
conc_c1 <- conc_c1 |>
  dplyr::bind_rows(
    tibble::tibble(id = unique(conc_c1$id), time = 0, Cc = 0)
  ) |>
  dplyr::distinct(id, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

dose_c1 <- cohort |>
  dplyr::transmute(id, time = 0, amt = teliso_dose(WT, 1.9), treatment = "1.9 mg/kg Q2W")

conc_c1$treatment <- "1.9 mg/kg Q2W"

o_conc <- PKNCA::PKNCAconc(conc_c1, Cc ~ time | id / treatment)
o_dose <- PKNCA::PKNCAdose(dose_c1, amt ~ time | id + treatment)
o_data <- PKNCA::PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(
    start = 0, end = tau,
    cmax = TRUE, tmax = TRUE, auclast = TRUE, cav = TRUE, half.life = TRUE
  )
)
res <- suppressWarnings(PKNCA::pk.nca(o_data))
nca <- as.data.frame(res)
```

```{r pknca-summary, echo = FALSE}
knitr::kable(
  nca |>
    dplyr::group_by(PPTESTCD) |>
    dplyr::summarise(
      Median = median(PPORRES, na.rm = TRUE),
      `P25`  = quantile(PPORRES, 0.25, na.rm = TRUE),
      `P75`  = quantile(PPORRES, 0.75, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::rename("NCA parameter" = PPTESTCD),
  digits = 3,
  caption = "Non-compartmental analysis of the simulated cycle-1 conjugate profiles at 1.9 mg/kg Q2W."
)
```

## Comparison against the published exposure distribution

Babel 2026 tabulates no conjugate NCA parameters, but the legend of Figure 3
prints the **cycle-1 average conjugate concentration (CavgC1) quartile
boundaries** of the 193-patient exposure-efficacy analysis set, in ug/mL:
Q1 `[2.20, 4.71]`, Q2 `[4.71, 6.17]`, Q3 `[6.17, 7.29]`, Q4 `[7.29, 10.8]`.
PKNCA's `cav` over the interval `[0, 14]` days is the same quantity, so the two
can be compared directly.

```{r check-cav}
cav_sim <- nca |>
  dplyr::filter(PPTESTCD == "cav") |>
  dplyr::pull(PPORRES)

cav_cmp <- tibble::tibble(
  Quantile  = c("25th (Q1/Q2 boundary)", "50th (Q2/Q3 boundary)", "75th (Q3/Q4 boundary)"),
  Simulated = as.numeric(quantile(cav_sim, c(0.25, 0.5, 0.75))),
  Published = c(4.71, 6.17, 7.29)
) |>
  dplyr::mutate(`Ratio` = Simulated / Published)

knitr::kable(cav_cmp, digits = 3,
             caption = "Simulated cycle-1 Cavg quartile boundaries against the Figure 3 legend.")

stopifnot(
  # Centre: a mis-transcribed CL, dose or unit would move the whole
  # distribution by tens of percent.
  abs(cav_cmp$Ratio[2] - 1) < 0.20,
  # Envelope, robust to which subjects land in the tails.
  all(abs(cav_cmp$Ratio - 1) < 0.30)
)
```

The simulated quartiles sit slightly **above** the published ones, and in a
direction the paper explains: its exposure metrics are computed "utilizing
actual doses received by patients" over the whole treatment period, whereas the
simulation gives every subject the full nominal dose at every cycle. The
published analysis set also mixes in 25 patients dosed at 1.6 mg/kg.

```{r nca-comparison, echo = FALSE}
knitr::kable(
  nlmixr2lib::ncaComparisonTable(
    simulated = tibble::tibble(cav = median(cav_sim)),
    reference = tibble::tibble(cav = 6.17),
    label_first_column = "NCA parameter"
  ),
  caption = "Median simulated cycle-1 Cavg against the Figure 3 published median (Q2/Q3 boundary)."
)
```

# Check 4 -- covariate effects reproduce the Figure 1 forest plot

For the conjugate, `AUCtau` depends on clearance only, so a purely categorical
covariate on CL implies an exposure ratio of exactly `1 / factor` when nothing
else changes. Figure 1A prints the simulated geometric-mean ratios, which do
include the covariate correlations of the real cohort, so exact agreement is not
expected -- but the sign and the order of magnitude are a real check.

```{r check-forest, echo = FALSE}
knitr::kable(
  tibble::tribble(
    ~Comparison,                      ~`Table S7 factor on CL`, ~`Implied AUCtau ratio`, ~`Figure 1A AUCtau ratio`,
    "ADA positive vs negative",        1.17,                     1 / 1.17,                0.846,
    "Asian vs White",                  0.887,                    1 / 0.887,               1.10,
    "Black or African American vs White", 0.763,                 1 / 0.763,               1.56
  ),
  digits = 3,
  caption = "Univariate implication of the Table S7 categorical CL factors against the Figure 1A simulated ratios."
)
```

The ADA and Asian rows agree closely (0.855 against 0.846; 1.127 against 1.10).
The Black or African American row does not (1.31 against 1.56), which is
expected: that stratum has only 8 patients, whose weights and albumin values
enter the simulated ratio alongside the race factor itself. Babel 2026 draws
the same conclusion, noting that those patients' exposures "were within the
variability of exposures in White patients" and that "the small number of Black
or African American patients (n = 8) limits conclusions on relevance".

# Unconjugated MMAE payload

The payload model is flip-flop: the deconjugation rate is much slower than MMAE
elimination, so the apparent terminal half-life of the payload is set by `ka`.

```{r check-mmae}
ka_typ  <- 0.154
cl_typ  <- 76.3
vc_typ  <- 96.0
kel_typ <- cl_typ / vc_typ

payload_halflives <- tibble::tibble(
  Quantity = c("deconjugation half-life = log(2)/ka",
               "MMAE elimination half-life = log(2)/(CL/Vc)"),
  Days     = c(log(2) / ka_typ, log(2) / kel_typ)
)
knitr::kable(payload_halflives, digits = 3,
             caption = "The payload system is flip-flop: input is slower than elimination.")

stopifnot(ka_typ < kel_typ)
```

Because Babel 2026 does not report the drug-antibody ratio, the molecular
weights, or the systemically available payload fraction, the depot of
`Babel_2026_telisotuzumab_mmae` must be dosed in **MMAE-equivalent mass**, not
in milligrams of conjugate. The vignette therefore solves the payload model per
unit MMAE-equivalent dose for the structural checks, and only afterwards
applies an illustrative conversion for plotting.

```{r sim-mmae}
rxode2::rxSetSeed(20260213L)
obs_t_m <- c(seq(0, 1, by = 0.02), seq(1.25, tau, by = 0.25))
ev_m <- dplyr::bind_rows(
  data.frame(id = cohort$id, time = 0, amt = 1, evid = 1L, cmt = "depot"),
  data.frame(
    id   = rep(cohort$id, each = length(obs_t_m)),
    time = rep(obs_t_m, times = n_sub),
    amt  = 0, evid = 0L, cmt = "central"
  )
) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  dplyr::left_join(cohort, by = "id")

sim_mmae <- as.data.frame(
  rxode2::rxSolve(mod_mmae, events = ev_m, returnType = "data.frame")
)
```

## Check 5 -- payload mass balance

```{r check-mmae-mass}
mb_m <- sim_mmae |>
  dplyr::group_by(id) |>
  dplyr::summarise(
    auc    = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    remain = dplyr::last(central) + dplyr::last(depot),
    cl_i   = dplyr::first(cl),
    .groups = "drop"
  ) |>
  dplyr::mutate(balance = (auc * cl_i + remain) / 1)

stopifnot(
  abs(median(mb_m$balance) - 1) < 0.01,
  quantile(abs(mb_m$balance - 1), 0.9) < 0.03
)
knitr::kable(
  tibble::tibble(
    Quantity = c("median (eliminated + remaining) / dose", "90th pct |deviation|"),
    Value    = c(median(mb_m$balance), quantile(abs(mb_m$balance - 1), 0.9))
  ),
  digits = 4,
  caption = "Mass balance of the payload model over 14 days, per unit MMAE-equivalent dose."
)
```

# Exposure-response

## How the four logistic models were recovered

Babel 2026 reports a nominal p value for each of the four relationships but
**tabulates no regression coefficients anywhere** -- not in the article, not in
Data S1. The only representation of the fitted models is the solid line in each
figure panel. The library admits a figure-derived parameter exactly when there
is no competing printed value, which is the case here.

Two properties of this paper make the digitisation sound rather than a guess:

1. **The plotted curve is the final model.** The usual objection to digitising
   an exposure-response panel is that the plotted line is the unadjusted,
   exposure-only fit while the tabulated model is covariate-adjusted. Babel
   2026 states that "no covariates were found to have a significant effect on
   efficacy or safety", so adjusted and unadjusted coincide for every endpoint.
2. **The functional form is decidable from the figure itself.** The Methods
   evaluated "linear and logarithmic logistic regression analyses"; fitting both
   to each digitised curve separates them decisively, and the winner differs
   between the conjugate endpoints and the payload endpoint.

```{r digitisation, echo = FALSE}
knitr::kable(
  tibble::tribble(
    ~Panel,                ~`Selected form`, ~`RMS residual, log form (pp)`, ~`RMS residual, linear form (pp)`,
    "Figure 2 (ORR)",       "logarithmic",    0.186, 0.703,
    "Figure 4 left (PN)",   "logarithmic",    0.083, 0.247,
    "Figure 4 right (CE)",  "logarithmic",    0.078, 0.441,
    "Figure 5 (TEAE)",      "linear",         7.598, 0.053
  ),
  digits = 3,
  caption = "Root-mean-square residual of each candidate form against the digitised curve, in percentage points of probability. The selected form is the smaller of the two in every panel."
)
```

The payload panel is additionally decidable by eye: its curve meets the left
axis at a finite 25.3%, whereas any logarithmic form is pinned to zero there.

## Replicating the published curves

```{r plot-er, echo = FALSE, fig.height = 3.4, fig.cap = "Replicates Figure 2 and Figure 4 of Babel 2026: fitted exposure-response curves against conjugate CavgADC, with the binned observed proportions read off the same panels."}
solve_prob <- function(model, output, cav) {
  ev <- data.frame(id = seq_along(cav), time = 0, amt = 0, evid = 0L, CAV = cav)
  as.data.frame(rxode2::rxSolve(model, events = ev, returnType = "data.frame"))[[output]]
}

cav_grid <- seq(0.2, 10, by = 0.1)
curves <- dplyr::bind_rows(
  tibble::tibble(CAV = cav_grid, p = solve_prob(mod_orr, "prob_orr_central", cav_grid),
                 Endpoint = "ORR per ICR (Figure 2)"),
  tibble::tibble(CAV = cav_grid,
                 p = solve_prob(mod_pn, "prob_peripheral_neuropathy_grade3", cav_grid),
                 Endpoint = "Grade >= 3 peripheral neuropathy (Figure 4)"),
  tibble::tibble(CAV = cav_grid,
                 p = solve_prob(mod_ce, "prob_corneal_epitheliopathy_grade2", cav_grid),
                 Endpoint = "Grade >= 2 corneal epitheliopathy (Figure 4)")
)

# Binned observed proportions read off the published panels (solid symbols).
observed <- tibble::tribble(
  ~Endpoint,                                     ~CAV, ~p,
  "ORR per ICR (Figure 2)",                       3.6, 0.11,
  "ORR per ICR (Figure 2)",                       5.1, 0.25,
  "ORR per ICR (Figure 2)",                       6.6, 0.295,
  "ORR per ICR (Figure 2)",                       8.5, 0.55,
  "Grade >= 3 peripheral neuropathy (Figure 4)",  3.6, 0.015,
  "Grade >= 3 peripheral neuropathy (Figure 4)",  5.1, 0.075,
  "Grade >= 3 peripheral neuropathy (Figure 4)",  6.4, 0.105,
  "Grade >= 3 peripheral neuropathy (Figure 4)",  8.2, 0.185,
  "Grade >= 2 corneal epitheliopathy (Figure 4)", 3.6, 0.015,
  "Grade >= 2 corneal epitheliopathy (Figure 4)", 5.1, 0.075,
  "Grade >= 2 corneal epitheliopathy (Figure 4)", 6.7, 0.12,
  "Grade >= 2 corneal epitheliopathy (Figure 4)", 8.4, 0.37
)

ggplot2::ggplot(curves, ggplot2::aes(CAV, p)) +
  ggplot2::geom_line(linewidth = 0.8, colour = "#26456e") +
  ggplot2::geom_point(data = observed, size = 1.6, colour = "#26456e") +
  ggplot2::facet_wrap(~Endpoint) +
  ggplot2::scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  ggplot2::labs(x = "CavgADC (ug/mL)", y = "Probability") +
  ggplot2::theme_bw()
```

```{r plot-teae, echo = FALSE, fig.height = 3.2, fig.cap = "Replicates the CavgMMAE panel of Figure 5 of Babel 2026."}
mmae_grid <- seq(0, 4, by = 0.05)
tibble::tibble(CAV = mmae_grid, p = solve_prob(mod_teae, "prob_teae_grade3", mmae_grid)) |>
  ggplot2::ggplot(ggplot2::aes(CAV, p)) +
  ggplot2::geom_line(linewidth = 0.8, colour = "#26456e") +
  ggplot2::geom_point(
    data = tibble::tibble(CAV = c(0.85, 1.35, 2.0, 3.0), p = c(0.42, 0.39, 0.59, 0.81)),
    size = 1.6, colour = "#26456e"
  ) +
  ggplot2::scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  ggplot2::labs(x = "CavgMMAE (ng/mL)", y = "Probability of a grade >= 3 TEAE") +
  ggplot2::theme_bw()
```

## Check 6 -- the exposure ratio implied by Table 1 equals the dose ratio

This is the sharpest available gate on the digitised **slopes**, and it does not
involve the intercepts at all. Table 1 prints the simulated median probability
of each endpoint at 1.6 and at 1.9 mg/kg Q2W. For a logistic in `log(Cavg)`,

`logit(p_1.9) - logit(p_1.6) = b * log(Cavg_1.9 / Cavg_1.6)`

so the printed probability pair and the digitised `b` together imply an exposure
ratio, which must equal the dose ratio 1.9/1.6 = 1.1875 because conjugate
clearance is linear.

```{r check-ratio}
logit <- function(p) log(p / (1 - p))

ratio_chk <- tibble::tribble(
  ~Endpoint,                          ~p16,    ~p19,    ~b,
  "ORR per ICR",                       0.222,   0.314,   2.450,
  "Grade >= 3 peripheral neuropathy",  0.0610,  0.0960,  2.793,
  "Grade >= 2 corneal epitheliopathy", 0.0780,  0.150,   4.206
) |>
  dplyr::mutate(
    `Implied exposure ratio` = exp((logit(p19) - logit(p16)) / b),
    `Dose ratio`             = 1.9 / 1.6,
    `Relative error`         = `Implied exposure ratio` / `Dose ratio` - 1
  )

knitr::kable(ratio_chk, digits = 4,
             caption = "Table 1 printed probabilities plus the digitised slope imply the dose ratio.")

stopifnot(all(abs(ratio_chk$`Relative error`) < 0.05))
```

All three land within 2.2% of the nominal dose ratio -- 0.6% and 0.2% for the
two safety endpoints -- on a quantity that was not used to fit the curves.

## Check 7 -- absolute probabilities against Table 1 at the published exposure

Check 6 gates the slopes without touching the intercepts; this one gates the
intercepts. Both sides are printed quantities: the exposure is the Figure 3
legend's median cycle-1 conjugate Cavg of 6.17 ug/mL at the LUMINOSITY dose
levels, scaled dose-proportionally to the 1.6 mg/kg arm (conjugate clearance is
linear, and non-linear clearance was tested and rejected), and the target is
Table 1.

```{r check-table1}
cav19_pub <- 6.17                      # Figure 3 legend, CavgC1 Q2/Q3 boundary
cav16_pub <- cav19_pub * 1.6 / 1.9     # linear clearance

table1 <- tibble::tribble(
  ~Endpoint,                           ~Model,  ~Output,                               ~p16_pub, ~p19_pub,
  "ORR per ICR",                       "orr",   "prob_orr_central",                     22.2,     31.4,
  "Grade >= 3 peripheral neuropathy",  "pn",    "prob_peripheral_neuropathy_grade3",     6.10,     9.60,
  "Grade >= 2 corneal epitheliopathy", "ce",    "prob_corneal_epitheliopathy_grade2",    7.80,    15.0
)
er_mods <- list(orr = mod_orr, pn = mod_pn, ce = mod_ce)

table1 <- table1 |>
  dplyr::rowwise() |>
  dplyr::mutate(
    p16_mod = 100 * solve_prob(er_mods[[Model]], Output, cav16_pub),
    p19_mod = 100 * solve_prob(er_mods[[Model]], Output, cav19_pub)
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(d16 = p16_mod - p16_pub, d19 = p19_mod - p19_pub)

knitr::kable(
  table1 |>
    dplyr::select(Endpoint, p16_pub, p16_mod, d16, p19_pub, p19_mod, d19) |>
    dplyr::rename(
      "Published 1.6 (%)" = p16_pub, "Model 1.6 (%)" = p16_mod,
      "Difference 1.6 (pp)" = d16,
      "Published 1.9 (%)" = p19_pub, "Model 1.9 (%)" = p19_mod,
      "Difference 1.9 (pp)" = d19
    ),
  digits = 2,
  caption = "Table 1 of Babel 2026 against the digitised logistic models evaluated at the Figure 3 published median exposure."
)

stopifnot(
  # Centre: a wrong intercept or a wrong exposure scale moves every row.
  abs(median(c(table1$d16, table1$d19))) < 3,
  # Envelope: the digitised intercepts carry a few percentage points of error.
  max(abs(c(table1$d16, table1$d19))) < 6
)
```

Two caveats keep this from being a tighter gate than it is. Table 1's exposure
metric is the whole-treatment Cavg, not the cycle-1 Cavg whose quartiles
Figure 3 prints, and the two differ by the accumulation ratio; and Table 1's
simulation resamples 500 patients from both studies while Figure 3's quartiles
describe the 193-patient efficacy set. The corneal row carries the largest
residual, consistent with the internal spread of the digitisation: solving each
endpoint's own intercept for the exposure that would reproduce its Table 1 value
gives 6.42 ug/mL from the neuropathy model and 6.78 ug/mL from the corneal
model, a 5.5% spread on two endpoints fitted to the *same* analysis set. That
spread is the practical accuracy of a figure-derived intercept.

## The simulated cohort's own exposure, for reference

```{r cav-ss, echo = FALSE}
cav_ss <- sim_adc |>
  dplyr::filter(time >= (n_dose - 1) * tau) |>
  dplyr::group_by(dose_mgkg, id) |>
  dplyr::summarise(
    cav = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2) / tau,
    .groups = "drop"
  ) |>
  dplyr::group_by(dose_mgkg) |>
  dplyr::summarise(
    `Median steady-state Cavg (ug/mL)` = median(cav),
    `P25` = quantile(cav, 0.25), `P75` = quantile(cav, 0.75),
    .groups = "drop"
  ) |>
  dplyr::rename("Dose (mg/kg Q2W)" = dose_mgkg)

knitr::kable(cav_ss, digits = 2,
             caption = "Steady-state conjugate Cavg of the simulated cohort, at nominal dosing.")
```

These sit above the published exposure used in Check 7, for the reason already
given under Check 4: the paper's metrics use the doses patients actually
received across the whole treatment period, and Babel 2026's own Discussion
notes that the positive exposure-safety relationships "are managed in the
clinical setting through dose modifications and/or dose reductions". A
simulation at full nominal dosing has no such attrition, so its exposure is an
upper bound on the observed one and must not be substituted into a
published-probability gate.

## Payload exposure and the TEAE model

```{r illustrative-payload}
# NOT a model parameter and NOT a published value. Babel 2026 never reports the
# drug-antibody ratio, the molecular weights, or the systemically available
# payload fraction, so this factor was back-solved from the CavgMMAE axis of
# Figure 5 purely so that the payload model can be plotted on a clinical dose.
# It is deliberately kept in the vignette and out of the model file.
payload_fraction_illustrative <- 0.0131

cav_mmae_ngml <-
  1000 * payload_fraction_illustrative * teliso_dose(68.9, 1.9) / (76.3 * tau)

teae_p <- 100 * solve_prob(mod_teae, "prob_teae_grade3", cav_mmae_ngml)
```

At the median weight of 68.9 kg and the illustrative payload fraction, the
1.9 mg/kg Q2W regimen gives a steady-state payload Cavg of
`r round(cav_mmae_ngml, 2)` ng/mL, inside the 0-4 ng/mL range of Figure 5, and
a predicted grade >= 3 TEAE probability of `r round(teae_p, 1)`%. Because the
fraction was calibrated to that same figure, the payload concentration **level**
is not an independent check; only the shape of the profile and the mass-balance
identity above are.

# Assumptions and deviations

* **The payload depot is dosed in MMAE-equivalent mass, not in milligrams of
  conjugate.** Babel 2026's Table S7 payload clearance (76.3 L/day) and volume
  (96.0 L) are physiologically sized for MMAE itself, but the paper reports
  neither the drug-antibody ratio nor the molecular weights nor the
  systemically available payload fraction, so the conversion from an
  administered conjugate dose cannot be reconstructed from anything on disk. The
  model file states this in its `description`, its `units$dosing` and its
  `compartmentData`. The vignette's illustrative fraction of 1.31% was
  back-solved from the Figure 5 CavgMMAE axis and is deliberately **not** part
  of the packaged model, following the precedent of
  `Wang_2024_omega3PUFA.Rmd`.
* **The payload CL-Vc random-effect correlation is not reproduced.** Babel 2026
  Results states that the payload model included "the inclusion of correlation
  between CL and Vc", but Table S7 tabulates only the three diagonal variances
  and no covariance or correlation coefficient. The omega block is therefore
  left diagonal rather than invented. Simulated payload variability is
  consequently slightly mis-shaped, although the marginal variances are correct.
* **The four exposure-response models carry digitised, not printed,
  coefficients.** See the Exposure-response section for the method, the
  form-selection residuals and the two independent cross-checks against Table 1.
  Treat the slopes as accurate to a few percent and the intercepts to a few
  percentage points of probability. Every `ini()` line and the model
  `description` carries this provenance.
* **Moderate and severe renal impairment share one estimated factor.** Babel
  2026 pooled the two strata (only 2 of 304 patients had severe impairment) and
  printed a single "Moderate/Severe vs. Normal" value. The model keeps
  `RENALIMP_MOD` and `RENALIMP_SEV` as separate columns so user data can retain
  the clinical distinction, and applies the one published factor to both.
* **Infusion duration is assumed to be 30 minutes.** Babel 2026 describes the
  route as "an intravenous infusion" and its sampling schedule references "the
  end of infusion" but never prints a duration. The conjugate model has no
  absorption step, so the assumption affects only the shape of the first few
  minutes of the peak and not AUC.
* **Treatment-emergent ADA status is carried as a subject-level indicator.**
  Nominally the covariate is time-varying, since a patient becomes ADA positive
  at seroconversion. Babel 2026 does not state whether it was implemented as a
  time-varying flag or an ever-positive subject-level flag, so the simpler
  reading is used.
* **The conjugate's additive residual row is printed as "Additional Error
  (Variance)".** The payload block of the same table prints "Additive Error
  (Variance)" for the identical quantity; the conjugate wording is read as a
  typographical slip and the value is used as an additive error.
* **Grade >= 2 peripheral neuropathy, all-grade and grade >= 2 ILD /
  pneumonitis, dose interruption or discontinuation due to an adverse event, DCR,
  and the DoR / PFS / OS Cox models are all reported as significant but are not
  plotted with a fitted curve**, so no coefficients are recoverable for them and
  they are not packaged. Figure 5's three additional payload metrics
  (CavgC1MMAE, CmingmMMAE, CmaxgmMMAE) are plotted, but they describe the same
  endpoint through a different exposure metric, so only the Cavg panel is
  packaged in order to keep the exposure metric aligned with the three
  conjugate-driven models.
* **The virtual cohort uses independent marginal covariate draws.** Table S4
  reports marginal distributions only; the real correlations between body
  weight, sex, albumin, age and renal function are not published, so the cohort
  cannot reproduce them. This matters for the Figure 1 forest comparison, which
  is why that section is presented descriptively rather than as a gate.
* **New canonical names introduced by this extraction.**
  `prob_peripheral_neuropathy_grade3` and
  `prob_corneal_epitheliopathy_grade2` are registered in
  `inst/references/compartment-names.md`, following the `prob_<endpoint>` shape
  founded by `prob_roc`. The four exposure-response models reuse the existing
  `CAV` covariate canonical, and their entries were added to that column's
  example-model list; note that this paper is the first in the register where a
  single `CAV` name carries two different analytes and two different units
  within one publication.
* **Residual error on the exposure-response models.** The source fits binomial
  logistic regressions with an exact Bernoulli likelihood and estimates no
  residual error. Each model emits its probability with a tiny fixed additive
  placeholder so that rxode2 accepts an observation declaration; this does not
  perturb the predicted probability. To simulate binary outcomes, apply
  `rbinom(n, 1, prob_*)` to the `rxSolve()` output.
