---
title: "MK-ASODN nanoliposomes, monkey and human PBPK (Bai 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MK-ASODN nanoliposomes, monkey and human PBPK (Bai 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Bai_2021_mkAsodnLiposome_human_pbpk"))$reference`
* Article: <https://doi.org/10.3389/fphar.2021.769538>

Monkey model:

```{r description-monkey, echo = FALSE, results = "asis"}
cat(rxode2::rxode(readModelDb("Bai_2021_mkAsodnLiposome_monkey_pbpk"))$description)
```

Human model:

```{r description-human, echo = FALSE, results = "asis"}
cat(rxode2::rxode(readModelDb("Bai_2021_mkAsodnLiposome_human_pbpk"))$description)
```

Bai 2021 studied MK-ASODN nanoliposomes, an antisense oligonucleotide against
midkine mRNA packaged in nanoliposomes for hepatocellular carcinoma. The
authors ran non-compartmental PK in monkeys, tissue distribution and excretion
in rats, and plasma protein binding across species. They then built a
whole-body physiologically based pharmacokinetic (PBPK) model of the monkey
data in GastroPlus 8.0 and extrapolated it to a human to predict exposure after
a 90 mg first-in-human dose.

The GastroPlus model has 14 perfusion-limited tissue compartments. Its organ
volumes and blood flows come from the software's built-in physiology, and its
tissue-to-plasma partition coefficients come from the built-in Rodgers-Single
(Lukacova) method. None of those values is printed, so the whole-body model
cannot be rebuilt from the paper. The paper *does* print enough to pin down
what that model does at the plasma level, and both files encode that
plasma-level model:

* **Monkey.** Table 4 gives predicted AUC0-inf on three dose arms. Dividing
  dose by AUC gives the same clearance on all three arms, 0.1002 L/h/kg, to
  five significant figures. The volume is not printed. It is recovered from
  the terminal slope of the simulated curves in Figure 1.
* **Human.** Results 3.6 prints CL = 4 L/h and Vss = 7.89 L directly.

The rat work produced tissue-distribution NCA only, with no model, so there
is nothing to extract from it.

# Population

Nine macaque monkeys (five male, four female, about 6 kg each) received a
single intravenous injection of 11.5, 23 or 46 mg/kg, three per dose (Methods
2.2 and 2.4). Plasma was sampled before dosing and at 0.08-6 h. The PBPK
simulation represents one typical monkey on GastroPlus default physiology.

The human model is a prediction for one GastroPlus virtual Chinese male. It
was not fitted to any human data. The 90 mg dose is the 46 mg/kg monkey dose
converted by body surface area with a safety factor of 10 (Methods 2.7).

# Source trace

| Quantity | Value | Source |
|---|---|---|
| Monkey `lcl` | log(0.1002 L/h/kg x 6 kg) | Table 4: dose / predicted AUC0-inf, identical on all three arms |
| Monkey `lvc` | log(0.1326 L/kg x 6 kg) | Figure 1A and 1B simulated curves: V = CL / k with k = 0.7556 /h (digitised) |
| Monkey reference weight 6 kg | 6 kg | Methods 2.2, "each weighing approximately 6 kg" |
| Monkey per-kg scaling | exponent 1 | Table 1 reports CL per kg; doses per kg |
| Human `lcl` | log(4 L/h) | Results 3.6 |
| Human `lvc` | log(7.89 L) | Results 3.6 (Vss) |
| `addSd`, `propSd` | fixed(0) | No variability reported |
| One-compartment IV structure | `d/dt(central) <- -kel * central` | Justified below from Table 4 and Figures 1 and 3 |
| Allometric exponent 0.8 (documentation only) | 0.8 | Methods 2.7 equation |

# Why a one-compartment reduction is the right shape

A reduction of a whole-body model is only useful if the paper's own outputs
show it is faithful. Four checks run on published numbers alone.

```{r reduction-checks}
# Table 4: PBPK-predicted AUC0-inf (ug.h/mL = mg.h/L) for the monkey arms.
dose_mgkg <- c(11.5, 23, 46)
auc_pred_monkey <- c(114.77, 229.54, 459.09)
cl_per_kg <- dose_mgkg / auc_pred_monkey
cl_per_kg

# 1. Linearity. The three arms give the same clearance to five significant
#    figures, so the PBPK clearance is dose-independent.
stopifnot(max(abs(cl_per_kg / 0.1002 - 1)) < 1e-4)

# 2. Human AUC identity: dose / CL against the Table 4 prediction.
auc_human_identity <- 90 / 4
c(identity = auc_human_identity, table4 = 22.48)
stopifnot(abs(auc_human_identity / 22.48 - 1) < 0.005)

# 3. Human half-life: ln2 * Vss / CL against the terminal slope digitised
#    by the maintainers from Figure 3 (log-linear regression over 0.5-6 h,
#    k = 0.4927 /h, t1/2 = 1.407 h).
t12_identity <- log(2) * 7.89 / 4
c(identity = t12_identity, figure3 = 1.407)
stopifnot(abs(t12_identity / 1.407 - 1) < 0.05)

# 4. Monkey volume: the digitised Figure 1 terminal rate constants of panels
#    A (0.7557 /h) and B (0.7555 /h) give V = CL / k. The back-extrapolated
#    intercept of each fit gives dose / C0 independently.
k_fig1 <- c(A = 0.7557, B = 0.7555)
v_from_k <- 0.1002 / k_fig1
v_from_c0 <- c(A = 0.1327, B = 0.1335)
rbind(v_from_k, v_from_c0)
stopifnot(max(abs(v_from_k / 0.1326 - 1)) < 0.005,
          max(abs(v_from_c0 / 0.1326 - 1)) < 0.01)
```

Every check here compares printed or digitised numbers, and each is
deterministic, so the tight bounds are appropriate. The same volume also
explains Table 4's *predicted* monkey Cmax. Dose / 0.1326 L/kg gives 86.7,
173.5 and 346.9 ug/mL against the tabulated 84.09, 172 and 341. For the
monkey, GastroPlus therefore reported the concentration at the start of the
terminal phase rather than the instantaneous spike, which is drawn in Figure
1 but lasts under about 10 minutes. The human Cmax of 49.98 ug/mL *is* that
spike (Figure 3). See Assumptions and deviations.

# Virtual cohort

The paper simulates one typical subject per scenario, and the models carry no
random effects. So the "cohort" is one deterministic subject per published
dose.

```{r cohort}
wt_monkey <- 6
t_monkey <- sort(unique(c(seq(0, 1, by = 0.02), seq(1, 12, by = 0.1))))
t_human <- sort(unique(c(seq(0, 1, by = 0.02), seq(1, 24, by = 0.1))))

ev_monkey <- dplyr::bind_rows(lapply(seq_along(dose_mgkg), function(i) {
  data.frame(
    id = i, time = c(0, t_monkey),
    evid = c(1L, rep(0L, length(t_monkey))),
    amt = c(dose_mgkg[i] * wt_monkey, rep(0, length(t_monkey))),
    cmt = "central", WT = wt_monkey,
    treatment = paste0(dose_mgkg[i], " mg/kg")
  )
}))

ev_human <- data.frame(
  id = 10L, time = c(0, t_human),
  evid = c(1L, rep(0L, length(t_human))),
  amt = c(90, rep(0, length(t_human))),
  cmt = "central", treatment = "Human 90 mg"
)
```

# Simulation

```{r simulate}
mod_monkey <- readModelDb("Bai_2021_mkAsodnLiposome_monkey_pbpk")
mod_human <- readModelDb("Bai_2021_mkAsodnLiposome_human_pbpk")

# No etas are declared, so the models are already deterministic.
sim_monkey <- rxode2::rxSolve(mod_monkey, events = ev_monkey,
                              keep = "treatment") |>
  as.data.frame()
sim_human <- rxode2::rxSolve(mod_human, events = ev_human,
                             keep = "treatment") |>
  as.data.frame()

# The explicit ODE must be solved as written, not replaced by linCmt().
stopifnot(is.null(rxode2::rxode(mod_monkey)$linCmt),
          is.null(rxode2::rxode(mod_human)$linCmt))

# Mass balance at the first sampled time: the concentration at t = 0 is
# dose / V (one-compartment bolus).
c0_monkey <- sim_monkey |> dplyr::filter(time == 0) |> dplyr::pull(Cc)
stopifnot(length(c0_monkey) == 3,
          max(abs(c0_monkey / (dose_mgkg / 0.1326) - 1)) < 1e-4)
```

# Replicate published figures

```{r figure-1, fig.cap = "Replicates the simulated (solid-line) curves of Figure 1 of Bai 2021: monkey plasma concentration after a single IV injection of 11.5, 23 and 46 mg/kg. The reduction omits the sub-10-minute initial spike that the whole-body model draws at t = 0."}
sim_monkey |>
  dplyr::filter(time <= 6) |>
  ggplot(aes(time, Cc, colour = treatment)) +
  geom_line() +
  scale_y_log10(limits = c(0.1, 1000)) +
  labs(x = "Time (h)", y = "MK-ASODN plasma concentration (ug/mL)",
       colour = "Dose") +
  theme_bw()
```

```{r figure-3, fig.cap = "Replicates Figure 3 of Bai 2021: predicted human plasma concentration after a single 90 mg IV injection."}
sim_human |>
  dplyr::filter(time <= 6) |>
  ggplot(aes(time, Cc)) +
  geom_line(colour = "blue") +
  scale_y_log10(limits = c(0.1, 100)) +
  labs(x = "Time (h)", y = "MK-ASODN plasma concentration (ug/mL)") +
  theme_bw()
```

# PKNCA validation

```{r pknca}
sim_all <- dplyr::bind_rows(sim_monkey, sim_human)

sim_nca <- sim_all |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time = 0 row per subject (IV bolus; rxode2 returns the post-dose
# concentration at t = 0, which is kept by distinct()).
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "mg/L", timeu = "h")

dose_df <- dplyr::bind_rows(ev_monkey, ev_human) |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "mg")

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
```

## Comparison against the paper's own PBPK predictions

The reduction's job is to reproduce what the whole-body model predicted, so
Table 4's *predicted* values are the reference.

```{r nca-compare}
published <- tibble::tribble(
  ~treatment,     ~cmax,  ~aucinf.obs,
  "11.5 mg/kg",   84.09,       114.77,
  "23 mg/kg",    172.00,       229.54,
  "46 mg/kg",    341.00,       459.09,
  "Human 90 mg",  49.98,        22.48
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "treatment",
  units         = c(cmax = "mg/L", aucinf.obs = "mg*h/L"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Reduced models vs the GastroPlus PBPK predictions in Table 4 of Bai 2021.",
    "* marks a difference above 20%. The only starred row is the human Cmax,",
    "which is the platform's instantaneous-bolus spike; see Assumptions and",
    "deviations."
  )
)
```

```{r nca-scorecard}
sim_wide <- as.data.frame(nca_res) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

score <- published |>
  dplyr::left_join(sim_wide, by = "treatment", suffix = c("_pub", "_sim")) |>
  dplyr::mutate(
    auc_ratio  = aucinf.obs_sim / aucinf.obs_pub,
    cmax_ratio = cmax_sim / cmax_pub
  )

# The join must have matched every published row.
stopifnot(nrow(score) == 4, !anyNA(score$auc_ratio), !anyNA(score$cmax_ratio))

score |>
  dplyr::select(treatment, aucinf.obs_pub, aucinf.obs_sim, auc_ratio,
                cmax_pub, cmax_sim, cmax_ratio, half.life) |>
  dplyr::rename(
    "Scenario" = treatment,
    "AUC published" = aucinf.obs_pub, "AUC simulated" = aucinf.obs_sim,
    "AUC ratio" = auc_ratio,
    "Cmax published" = cmax_pub, "Cmax simulated" = cmax_sim,
    "Cmax ratio" = cmax_ratio, "t1/2 simulated (h)" = half.life
  ) |>
  knitr::kable(digits = 3,
               caption = "Reduced model / published PBPK prediction, per scenario.")

# Deterministic comparisons, so tight bounds are correct.
# AUC reproduces the PBPK prediction on every scenario.
stopifnot(max(abs(score$auc_ratio - 1)) < 0.01)

# Monkey Cmax reproduces the tabulated prediction (start of the terminal
# phase) on all three arms.
monkey <- score |> dplyr::filter(treatment != "Human 90 mg")
stopifnot(max(abs(monkey$cmax_ratio - 1)) < 0.05)

# Human Cmax: the reduction starts at dose / Vss = 11.4 ug/mL, i.e. about
# 0.23 of the platform's bolus spike. Pinned so a change in structure or
# volume shows up here.
human <- score |> dplyr::filter(treatment == "Human 90 mg")
stopifnot(abs(human$cmax_sim - 90 / 7.89) < 0.01,
          abs(human$cmax_ratio - 0.228) < 0.005)

# Half-lives: monkey 0.917 h from the digitised Figure 1 slope, human
# ln2 * 7.89 / 4 = 1.367 h.
stopifnot(max(abs(monkey$half.life / (log(2) * 0.1326 / 0.1002) - 1)) < 0.01,
          abs(human$half.life / t12_identity - 1) < 0.01)
```

## Comparison against the observed monkey data

The monkey NCA in Table 1 is observed data, which the whole-body model itself
fits only approximately. The paper's own fold errors are 2.22, 2.00 and 1.73
on Cmax and 0.94, 0.98 and 1.10 on AUC (Table 4). The reduction inherits
those, because its AUC equals the PBPK prediction.

```{r observed-monkey}
observed <- tibble::tribble(
  ~treatment,    ~auc_obs, ~cmax_obs, ~t12_obs, ~cl_obs_mlkgh, ~vss_obs_mlkg,
  "11.5 mg/kg",    122.04,    187.16,     0.79,         95.41,         88.41,
  "23 mg/kg",      232.87,    347.43,     1.08,         99.16,        104.63,
  "46 mg/kg",      415.38,    588.73,     1.33,        110.95,        137.89
)

obs_cmp <- observed |>
  dplyr::left_join(sim_wide, by = "treatment") |>
  dplyr::mutate(auc_sim_over_obs = aucinf.obs / auc_obs,
                cmax_sim_over_obs = cmax / cmax_obs)
stopifnot(nrow(obs_cmp) == 3, !anyNA(obs_cmp$auc_sim_over_obs))

obs_cmp |>
  dplyr::select(treatment, auc_obs, aucinf.obs, auc_sim_over_obs,
                cmax_obs, cmax, cmax_sim_over_obs, t12_obs, half.life) |>
  dplyr::rename(
    "Dose" = treatment,
    "AUC observed" = auc_obs, "AUC simulated" = aucinf.obs,
    "AUC sim/obs" = auc_sim_over_obs,
    "Cmax observed" = cmax_obs, "Cmax simulated" = cmax,
    "Cmax sim/obs" = cmax_sim_over_obs,
    "t1/2 observed (h)" = t12_obs, "t1/2 simulated (h)" = half.life
  ) |>
  knitr::kable(digits = 3,
               caption = "Reduced monkey model vs observed monkey NCA (Table 1 of Bai 2021).")

# The paper's own AUC fold errors (0.94-1.10) carried over.
stopifnot(all(obs_cmp$auc_sim_over_obs > 0.9 & obs_cmp$auc_sim_over_obs < 1.15))
```

The observed monkey data are less than dose-proportional (the observed
clearance rises from 95 to 111 mL/h/kg and the half-life from 0.79 to 1.33 h
across the dose range; Table 2 calls dose proportionality "inconclusive"). The
PBPK model, and therefore both reductions, is linear and does not reproduce
that trend.

# Assumptions and deviations

* **The whole-body structure is not reproduced.** The 14-tissue GastroPlus
  model uses the software's default monkey and human physiology and
  Rodgers-Single Kp values, none of which is printed. Both files encode the
  plasma-level behaviour the paper's own outputs pin down, not the platform
  model.
* **Monkey clearance is back-solved, not printed.** Methods 2.7 says the
  monkey PBPK used the in vivo clearance but does not say which value. Dose /
  predicted AUC0-inf gives 0.10020 L/h/kg on all three arms. That is close to,
  but not equal to, the mean of the three observed NCA clearances (101.8
  mL/h/kg, Table 1).
* **Monkey volume is digitised.** No monkey model volume is printed. The
  maintainers digitised the simulated curves of Figure 1 by pixel regression
  over the terminal phase (1.7-5.9 h). Panels A and B give V = CL / k = 0.1326
  L/kg to four digits. Panel C, on a coarser axis, gives 0.1341. The reduced
  model then reproduces Table 4's predicted AUC (by construction) and
  predicted Cmax (independently, within 3.1%).
* **Monkey per-kg scaling is linear, referenced to 6 kg.** The paper reports
  clearance per kg and doses per kg, so clearance and volume scale with body
  weight to the power 1. The 6 kg reference is the approximate weight stated in
  Methods 2.2. Under mg/kg dosing, concentrations do not depend on the weight
  chosen.
* **The initial spike is not carried.** Both GastroPlus curves (Figures 1 and
  3) start with a sub-10-minute spike, from the bolus mixing into a small
  plasma volume before tissue distribution. It holds under 2% of the AUC. The
  human Table 4 Cmax (49.98 ug/mL) is that spike, so the reduction gives 11.4
  ug/mL (ratio 0.23). The monkey Table 4 Cmax values are *not* the spike (they
  equal dose / V at the start of the terminal phase), so the reduction
  reproduces them. A two-compartment fit to the spike would need values the
  paper does not print. The bolus Cmax is therefore a known limitation of the
  human file: do not use it to predict a peak concentration within the first
  few minutes after an injection.
* **The human prediction is a single virtual subject.** The paper prints human
  CL and Vss but neither body weight in its allometric equation
  CL_human = CL_monkey x (BW_human / BW_monkey)^0.8. With the monkey
  clearance above (0.6012 L/h at 6 kg), CL_human = 4 L/h implies a human
  weight of about 64 kg. That is arithmetic, not a printed value, so the human
  file carries no weight covariate.
* **No variability.** The paper reports no interindividual or residual
  variability for either simulation. No etas are declared, and `addSd` /
  `propSd` are `fixed(0)`.
* **Protein binding.** Human plasma protein binding was 96-99% (Table 3). The
  paper reports only total concentrations, so no unbound concentration is
  computed.
* **Errata.** No erratum or correction for this article was found in a check
  of Europe PMC (PMID 34803711; no linked correction, and no erratum
  matching the article) on 2026-09-29.
