---
title: "Imipenem (Bai 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Imipenem (Bai 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r libraries, message = FALSE}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(ggplot2)
```

Bai J, Wen A, Li Z, Li X, Duan M (2024). Population pharmacokinetics and dosing
optimisation of imipenem in critically ill patients. *Eur J Hosp Pharm*
31(5):434-439. [doi:10.1136/ejhpharm-2022-003403](https://doi.org/10.1136/ejhpharm-2022-003403).
PMCID PMC11347199.

# Population

A prospective, open-label, single-centre study in the intensive care unit of
Beijing Friendship Hospital, Capital Medical University (Beijing, China).
Fifty-one critically ill adults meeting the Sepsis 3.0 diagnostic criteria and
treated with imipenem-cilastatin contributed 196 plasma imipenem
concentrations (3.84 samples per patient); 7 patients were sampled at all nine
nominal time points and 44 contributed 3-4 points. Sampling began at least 24
hours after the start of therapy, so all data are at or near steady state.

Patients were 18-96 years old (median 56); 18 of 51 (35.3%) were female.
Comorbid state at enrolment was severe: acute kidney injury in 39.2%, hepatic
insufficiency (Child-Pugh B or C) in 51.0%, septic shock in 35.3% and
mechanical ventilation in 51.0%. Creatinine clearance, computed with a
Cockcroft-Gault equation written in SI creatinine units, was 104.59 +/- 60.95
mL/min (median 99.34, range 17.80-256.22).

Imipenem was given intravenously as 0.5 g or 1 g diluted into 100 mL and
infused by pump over 1 hour; 0.5 g q6h, 0.5 g q8h and 1 g q8h were the
commonly prescribed regimens. Imipenem was assayed by HPLC-UV at 298 nm
(LLOQ 0.3 ug/mL, linear over 0.3-200.0 ug/mL) in MOPS-stabilised plasma.

The model was fit in Phoenix NLME 8.0 by first-order conditional
estimation-least squares (FOCE-ELS) and evaluated by goodness-of-fit plots, a
1000-replicate visual predictive check and a 1000-sample bootstrap.

```{r population}
mod <- nlmixr2lib::modellib("Bai_2024_imipenem")
str(rxode2::modelExtract(mod, "population"), max.level = 1)
```

# Source trace

Every structural equation and every `ini()` value, with the location in
Bai 2024 it was taken from.

```{r source_trace, echo = FALSE}
tibble::tribble(
  ~Item,                     ~Value,         ~`Source location`,
  "Structural model",        "2-compartment IV, zero-order infusion, first-order elimination", "Methods, 'Base model', Eqs. 3-6",
  "Central mass balance",    "dXc/dt = K0 - Q*(Cc - Cp) - CL*Cc; Xc(0) = 0", "Eq. 3",
  "Peripheral mass balance", "dXp/dt = Q*(Cc - Cp); Xp(0) = 0", "Eq. 4 (printed as 'Xc,0 = 0'; see Errata)",
  "Central concentration",   "Cc = Xc / Vc", "Eq. 5",
  "Peripheral concentration","Cp = Xp / Vp", "Eq. 6",
  "IIV model",               "Pi = P * exp(eta_i), eta_i ~ N(0, omega^2)", "Methods, 'Base model', Eq. 1",
  "Covariate model (general)", "Pi = P * (COV / COV_median)^f * exp(eta_i)", "Methods, 'Final model', Eq. 7",
  "Covariate model (final)", "CL (L/h) = 11.357 * (CrCl / 99.896)^0.473 * exp(eta)", "Results, 'Population pharmacokinetics', Eq. 8",
  "Residual error",          "Ci = C * (1 + eps), proportional", "Methods, 'Base model', Eq. 2",
  "CL = 11.357 L/h",         "RSE 3.024%; 95% CI 10.679-12.035", "Table 2, row 'CL (L/h)'",
  "Vc = 16.378 L",           "RSE 2.283%; 95% CI 15.641-17.116", "Table 2, row 'Vc (L)'",
  "Vp = 10.904 L",           "RSE 3.951%; 95% CI 10.054-11.754", "Table 2, row 'Vp (L)'",
  "Q = 7.645 L/h",           "RSE 4.086%; 95% CI 7.029-8.261", "Table 2, row 'Q (L/h)'",
  "e_crcl_cl = 0.473",       "RSE 5.181%; 95% CI 0.425-0.521", "Table 2, row 'fCrCl'; Eq. 8",
  "CrCl centring = 99.896 mL/min", "stated median of CrCl", "Text immediately below Eq. 8",
  "IIV Vc, CV 89.853%",      "omega^2 = log(0.89853^2 + 1) = 0.5918651", "Table 2, 'Iiv (CV%)' column",
  "IIV Vp, CV 8.319%",       "omega^2 = log(0.08319^2 + 1) = 0.0068967", "Table 2, 'Iiv (CV%)' column",
  "IIV Q, CV 24.453%",       "omega^2 = log(0.24453^2 + 1) = 0.0580754", "Table 2, 'Iiv (CV%)' column",
  "IIV CL, CV 35.748%",      "omega^2 = log(0.35748^2 + 1) = 0.1202617", "Table 2, 'Iiv (CV%)' column; restated in the Table 2 footnote",
  "propSd = 0.30370",        "RSE 5.126%; 95% CI 27.298-33.442", "Table 2, 'Residual error (proportional error, CV%)'"
) |>
  knitr::kable(caption = "Source trace for Bai 2024 imipenem.")
```

# Published parameter values, transcribed independently

Every check below is built from these literals, read off Bai 2024 Table 2 and
Eq. 8 and typed here by hand. They are deliberately **not** read back out of
the model object, so a mis-transcription in the model file makes the checks
fail rather than cancel out.

```{r published_values}
pub <- list(
  CL        = 11.357,  # Table 2, 'CL (L/h)'
  Vc        = 16.378,  # Table 2, 'Vc (L)'
  Vp        = 10.904,  # Table 2, 'Vp (L)'
  Q         =  7.645,  # Table 2, 'Q (L/h)'
  f_crcl    =  0.473,  # Table 2, 'fCrCl'; Eq. 8 exponent
  crcl_ref  = 99.896,  # text below Eq. 8
  prop_cv   =  0.30370 # Table 2, residual error
)
```

# Covariate model: reproducing Eq. 8

`cl` is returned by `rxSolve()` as a derived quantity, so it can be compared
directly against Eq. 8 evaluated by hand. The four evaluation points are the
midpoints of the CrCl bands Bai 2024 used to stratify its dosing simulations
(Results, 'Simulation': 17.80-30, 30-60, 60-90 and 90-256.22 mL/min), plus the
centring value itself.

```{r covariate_check}
crcl_check <- c(23.9, 45, 75, 173.1, pub$crcl_ref)

ev_point <- rxode2::et(amt = 1000, cmt = "central", dur = 1) |>
  rxode2::et(1, cmt = "central")

cl_model <- vapply(crcl_check, function(cc) {
  d <- as.data.frame(ev_point)
  d$CRCL <- cc
  s <- rxode2::rxSolve(rxode2::zeroRe(mod), d, returnType = "data.frame")
  s$cl[1]
}, numeric(1))

cl_paper <- pub$CL * (crcl_check / pub$crcl_ref)^pub$f_crcl

cov_tab <- tibble::tibble(
  `CrCl (mL/min)`      = crcl_check,
  `Eq. 8 CL (L/h)`     = round(cl_paper, 4),
  `Model CL (L/h)`     = round(cl_model, 4),
  `Relative difference` = signif(cl_model / cl_paper - 1, 3)
)
knitr::kable(cov_tab, caption = "Model clearance against Eq. 8 evaluated by hand.")

# Deterministic identity: a wrong exponent, a wrong centring constant or a
# wrong typical CL all break this.
stopifnot(max(abs(cl_model / cl_paper - 1)) < 1e-8)

# At the centring value the covariate term must be exactly 1, so CL must equal
# the tabulated typical value.
stopifnot(abs(cl_model[length(cl_model)] - pub$CL) < 1e-8)
```

# Structural check against an independent closed form

A two-compartment model with a zero-order input has a closed-form solution.
The helper below builds it from the published `CL`, `Vc`, `Vp` and `Q` only --
it never touches the model object or the ODE solver -- and steady state is
reached by superposing single-dose profiles. Comparing it against `rxSolve()`
on the typical subject (`zeroRe()`) tests that Eqs. 3-6 were translated into
`d/dt()` correctly.

```{r closed_form}
# Macro-constants for a 2-compartment model, from micro-constants.
biexp_coefs <- function(CL, Vc, Vp, Q) {
  k10 <- CL / Vc
  k12 <- Q / Vc
  k21 <- Q / Vp
  b <- k10 + k12 + k21
  disc <- sqrt(b^2 - 4 * k21 * k10)
  alpha <- (b + disc) / 2
  beta <- (b - disc) / 2
  list(
    alpha = alpha,
    beta = beta,
    A = (alpha - k21) / (Vc * (alpha - beta)),
    B = (k21 - beta) / (Vc * (alpha - beta))
  )
}

# Concentration after ONE zero-order infusion of `dose` given over `tinf`.
conc_single_infusion <- function(t, dose, tinf, cf) {
  k0 <- dose / tinf
  ton <- pmin(t, tinf)
  toff <- pmax(0, t - tinf)
  k0 * (cf$A / cf$alpha * (1 - exp(-cf$alpha * ton)) * exp(-cf$alpha * toff) +
        cf$B / cf$beta  * (1 - exp(-cf$beta  * ton)) * exp(-cf$beta  * toff))
}

# Steady state by superposition of `ndose` identical infusions given at
# 0, tau, 2*tau, ...  `t` is ABSOLUTE time from the first dose, not time
# within the interval.
conc_multiple_infusion <- function(t, dose, tinf, tau, ndose, cf) {
  out <- numeric(length(t))
  for (i in seq_len(ndose)) {
    td <- t - (i - 1) * tau
    out <- out + ifelse(td > 0, conc_single_infusion(pmax(td, 0), dose, tinf, cf), 0)
  }
  out
}

cf <- biexp_coefs(pub$CL, pub$Vc, pub$Vp, pub$Q)
c(alpha_half_life = log(2) / cf$alpha, beta_half_life = log(2) / cf$beta)
```

```{r closed_form_compare}
tau_ss <- 8      # 1 g q8h, one of the commonly prescribed regimens
tinf <- 1        # "administered by an intravenous infusion pump over 1 hour"
ndose <- 30      # 240 h of dosing; > 100 terminal half-lives

ev_ss <- rxode2::et(
  amt = 1000, cmt = "central", dur = tinf, ii = tau_ss, addl = ndose - 1
) |>
  rxode2::et(seq((ndose - 1) * tau_ss, ndose * tau_ss, by = 0.05), cmt = "central")

d_typ <- as.data.frame(ev_ss)
d_typ$CRCL <- pub$crcl_ref

typ <- rxode2::rxSolve(rxode2::zeroRe(mod), d_typ, returnType = "data.frame") |>
  dplyr::filter(!is.na(Cc), time >= (ndose - 1) * tau_ss)

typ$closed_form <- conc_multiple_infusion(
  typ$time, 1000, tinf, tau_ss, ndose, cf
)

rel_err <- with(typ, abs(Cc - closed_form) / closed_form)
c(max_relative_error = max(rel_err))

# Pure numerical error between two exact representations of the same system,
# so a tight bound is correct here (both sides use the same drawn parameters;
# see CLAUDE.md on when tight all() bounds apply).
stopifnot(max(rel_err) < 1e-4)
```

```{r closed_form_plot, echo = FALSE}
ggplot2::ggplot(typ, ggplot2::aes(x = time - (ndose - 1) * tau_ss)) +
  ggplot2::geom_line(ggplot2::aes(y = closed_form, colour = "Closed form"), linewidth = 1.6) +
  ggplot2::geom_line(ggplot2::aes(y = Cc, colour = "rxode2 solve"), linewidth = 0.6) +
  ggplot2::scale_colour_manual(values = c("Closed form" = "grey70", "rxode2 solve" = "black")) +
  ggplot2::labs(
    x = "Time within the dosing interval (h)",
    y = "Imipenem concentration (mg/L)",
    colour = NULL,
    title = "Typical subject, 1 g q8h as a 1-hour infusion, steady state",
    subtitle = "CrCl 99.896 mL/min (the Eq. 8 centring value)"
  ) +
  ggplot2::theme_bw()
```

# PKNCA validation

Bai 2024 publishes no non-compartmental analysis table, so the reference side
of the comparison below is derived from the paper's own printed parameters:

* `AUC0-tau` at steady state must equal `Dose / CL` by mass balance
  (`1000 / 11.357 = `r round(1000 / pub$CL, 3)` mg*h/L`). This is an identity
  of the published clearance, independent of everything else in the model.
* `Cav` over the interval is that AUC divided by `tau`.
* `Cmax`, `Cmin` and the terminal half-life come from the independent
  closed-form solution built above out of the published `Vc`, `Vp` and `Q`.

```{r pknca}
conc_df <- typ |>
  dplyr::transmute(
    id = 1L,
    treatment = "1 g q8h",
    time = time,
    Cc = Cc
  ) |>
  dplyr::filter(!is.na(Cc))

dose_df <- data.frame(
  id = 1L,
  treatment = "1 g q8h",
  time = (ndose - 1) * tau_ss,
  amt = 1000
)

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | id / treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | id + treatment)

intervals <- data.frame(
  start = (ndose - 1) * tau_ss,
  end = ndose * tau_ss,
  cmax = TRUE,
  cmin = TRUE,
  tmax = TRUE,
  auclast = TRUE,
  cav = TRUE,
  half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r nca_compare}
cf_grid <- seq((ndose - 1) * tau_ss, ndose * tau_ss, by = 1e-3)
cf_conc <- conc_multiple_infusion(cf_grid, 1000, tinf, tau_ss, ndose, cf)

published <- tibble::tibble(
  treatment = "1 g q8h",
  auclast   = 1000 / pub$CL,             # mass balance on the published CL
  cav       = 1000 / (pub$CL * tau_ss),
  cmax      = max(cf_conc),              # closed form from published Vc/Vp/Q/CL
  cmin      = min(cf_conc),
  tmax      = cf_grid[which.max(cf_conc)] - (ndose - 1) * tau_ss,
  half.life = log(2) / cf$beta
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "treatment",
  units = c(
    auclast = "mg*h/L", cav = "mg/L", cmax = "mg/L",
    cmin = "mg/L", tmax = "h", half.life = "h"
  ),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Steady-state NCA of the typical-subject simulation against values derived",
    "from Bai 2024 Table 2. * differs from reference by >20%."
  ),
  align = c("l", "l", "r", "r", "r")
)
```

```{r nca_gate}
nca_wide <- nca_res$result |>
  dplyr::select(PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

# AUC0-tau at steady state is Dose / CL exactly. A mis-transcribed CL, a
# missed covariate centring, or an infusion that did not deliver the full dose
# all break this.
stopifnot(abs(nca_wide$auclast / (1000 / pub$CL) - 1) < 0.005)

# Cmax and Cmin come from the distribution parameters, so these bound Vc, Vp
# and Q. Trapezoidal Cmax on a 0.05 h grid is slightly conservative.
stopifnot(
  abs(nca_wide$cmax / published$cmax - 1) < 0.01,
  abs(nca_wide$cmin / published$cmin - 1) < 0.01,
  abs(nca_wide$half.life / published$half.life - 1) < 0.02
)
```

# Replicating the paper's dosing recommendation (Figure 4)

Bai 2024 ran Monte Carlo simulations of six regimens -- 500 and 1000 mg as
1-hour infusions, each q6h, q8h and q12h -- and scored them on the fraction of
the dosing interval with concentration above the MIC (`fT>MIC`), taking 100%
`fT>MIC` as the target for critically ill patients. The Discussion states the
conclusion explicitly:

> If the MIC=1, to achieve a target of 100% *f*T>MIC or *f*T>4-5 x MIC, we
> administered a dosage schedule of 0.5 g q6h, 1 g q8h or 1 g q6h.

Because 100% `fT>MIC` is attained exactly when the steady-state trough exceeds
the MIC, this is a discrete, falsifiable claim: at MIC = 1 mg/L, exactly three
of the six regimens should clear the bar and three should not. The check below
evaluates all six on the typical subject at the Eq. 8 centring CrCl.

```{r regimen_grid}
regimens <- tibble::tibble(
  dose = c(500, 500, 500, 1000, 1000, 1000),
  tau  = c(6, 8, 12, 6, 8, 12)
) |>
  dplyr::mutate(
    treatment = sprintf("%s g q%dh", ifelse(dose == 500, "0.5", "1"), tau)
  )

solve_typical_regimen <- function(dose, tau, crcl) {
  nd <- ceiling(240 / tau)
  ev <- rxode2::et(amt = dose, cmt = "central", dur = tinf, ii = tau, addl = nd - 1) |>
    rxode2::et(seq((nd - 1) * tau, nd * tau, by = 0.02), cmt = "central")
  d <- as.data.frame(ev)
  d$CRCL <- crcl
  rxode2::rxSolve(rxode2::zeroRe(mod), d, returnType = "data.frame") |>
    dplyr::filter(!is.na(Cc), time >= (nd - 1) * tau)
}

typ_reg <- regimens |>
  dplyr::rowwise() |>
  dplyr::mutate(
    cmin = min(solve_typical_regimen(dose, tau, pub$crcl_ref)$Cc)
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    `100% fT>MIC at MIC = 1 mg/L` = ifelse(cmin > 1, "attained", "not attained")
  )

typ_reg |>
  dplyr::select(
    Regimen = treatment,
    `Steady-state Cmin (mg/L)` = cmin,
    `100% fT>MIC at MIC = 1 mg/L`
  ) |>
  dplyr::mutate(`Steady-state Cmin (mg/L)` = round(`Steady-state Cmin (mg/L)`, 3)) |>
  knitr::kable(
    caption = paste(
      "Typical subject at CrCl 99.896 mL/min. Bai 2024 Discussion names",
      "0.5 g q6h, 1 g q6h and 1 g q8h as the regimens attaining 100% fT>MIC",
      "at MIC = 1 mg/L."
    )
  )
```

```{r regimen_gate}
attained <- sort(typ_reg$treatment[typ_reg$cmin > 1])
paper_says <- sort(c("0.5 g q6h", "1 g q6h", "1 g q8h"))

attained
paper_says

# Exact set equality: 6 of 6 regimens classified as the paper classified them.
# Deterministic (typical value, no random draw), so this is machine-stable.
stopifnot(identical(attained, paper_says))
```

All six regimens are classified exactly as Bai 2024 classified them. The three
that fail do so with room to spare -- the largest failing trough is
`r round(max(typ_reg$cmin[typ_reg$cmin <= 1]), 3)` mg/L and the smallest
passing trough is `r round(min(typ_reg$cmin[typ_reg$cmin > 1]), 3)` mg/L --
so the agreement is not an artefact of a borderline call.

# Probability of target attainment by renal-function stratum

Bai 2024 Figure 4 gives PTA against MIC for the six regimens within each of
four CrCl strata (17.80-30, 30-60, 60-90 and 90-256.22 mL/min). The cohort
below reproduces that figure.

The paper does not report the within-stratum CrCl distributions it sampled, so
the cohort is drawn from a log-normal matched to the Table 1 marginal moments
(mean 104.59, SD 60.95 mL/min), truncated to the observed 17.80-256.22 range
and then split at the paper's stratum boundaries. See Errata.

```{r cohort}
n_per_arm <- 200L   # cap per the skill's 200-per-arm limit

set.seed(20240101L)
sigma2 <- log(1 + (60.95 / 104.59)^2)
mu_log <- log(104.59) - sigma2 / 2
pool <- rlnorm(2e5, mu_log, sqrt(sigma2))
pool <- pool[pool >= 17.80 & pool <= 256.22]

strata <- tibble::tribble(
  ~stratum,             ~lo,    ~hi,
  "CrCl 17.8-30",       17.80,  30,
  "CrCl 30-60",         30,     60,
  "CrCl 60-90",         60,     90,
  "CrCl 90-256"    ,    90,     256.22
)

c(cohort_mean = mean(pool), cohort_median = median(pool), cohort_sd = sd(pool))
```

```{r pta_simulation}
mic_grid <- 2^(-4:5)   # 0.0625 to 32 mg/L, the range in Bai 2024 Figure 4

simulate_arm <- function(stratum, lo, hi, dose, tau, treatment) {
  nd <- ceiling(240 / tau)
  base <- as.data.frame(
    rxode2::et(amt = dose, cmt = "central", dur = tinf, ii = tau, addl = nd - 1) |>
      rxode2::et(seq((nd - 1) * tau, nd * tau, by = 0.05), cmt = "central")
  )

  # Reseed inside the arm loop so every arm sees the same subjects: common
  # random numbers make the between-regimen comparison exact rather than noisy.
  set.seed(20240101L)
  in_band <- pool[pool >= lo & pool < hi]
  crcl <- sample(in_band, n_per_arm, replace = TRUE)

  d <- base[rep(seq_len(nrow(base)), times = n_per_arm), ]
  d$id <- rep(seq_len(n_per_arm), each = nrow(base))
  d$CRCL <- rep(crcl, each = nrow(base))

  rxode2::rxSetSeed(20240101L)
  s <- rxode2::rxSolve(mod, d, returnType = "data.frame") |>
    dplyr::filter(!is.na(Cc), time >= (nd - 1) * tau)

  cmin <- vapply(split(s$Cc, s$id), min, numeric(1))

  tibble::tibble(
    stratum = stratum,
    treatment = treatment,
    mic = mic_grid,
    pta = vapply(mic_grid, function(m) 100 * mean(cmin > m), numeric(1))
  )
}

arms <- tidyr::expand_grid(strata, regimens)

pta <- purrr::pmap_dfr(
  list(arms$stratum, arms$lo, arms$hi, arms$dose, arms$tau, arms$treatment),
  simulate_arm
) |>
  dplyr::mutate(
    stratum = factor(stratum, levels = strata$stratum),
    treatment = factor(treatment, levels = regimens$treatment)
  )
```

```{r pta_plot, fig.height = 6}
ggplot2::ggplot(pta, ggplot2::aes(x = mic, y = pta, colour = treatment)) +
  ggplot2::geom_hline(yintercept = c(80, 90), linetype = "dashed", colour = "grey60") +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::geom_point(size = 1) +
  ggplot2::scale_x_continuous(trans = "log2", breaks = mic_grid) +
  ggplot2::facet_wrap(~stratum) +
  ggplot2::labs(
    x = "MIC (mg/L)",
    y = "PTA for 100% fT>MIC (%)",
    colour = "Regimen",
    title = "Replicates Figure 4 of Bai 2024",
    subtitle = paste(
      "Percentage of", n_per_arm,
      "simulated patients per arm attaining 100% fT>MIC at steady state"
    )
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

```{r pta_table}
pta |>
  dplyr::filter(mic %in% c(0.5, 1, 2, 4)) |>
  tidyr::pivot_wider(names_from = mic, values_from = pta, names_prefix = "MIC ") |>
  dplyr::rename(Stratum = stratum, Regimen = treatment) |>
  knitr::kable(
    caption = "PTA (%) for 100% fT>MIC at the four clinically relevant MICs.",
    digits = 1
  )
```

```{r pta_gate}
by_stratum <- pta |>
  tidyr::pivot_wider(names_from = stratum, values_from = pta)

# 1. Renal-function ordering, at EVERY regimen and EVERY MIC. This is an exact
#    identity, not a statistical tendency: the four strata are disjoint and
#    ordered, every arm is solved from the same seed with the same number of
#    subjects (so subject j carries the same etas in all four strata), and
#    the Eq. 8 exponent is positive. Subject j therefore has a strictly lower
#    CL -- and so a strictly higher trough -- in a lower stratum. A sign error
#    on e_crcl_cl, or a covariate wired to the wrong parameter, inverts it.
stopifnot(
  all(by_stratum$`CrCl 17.8-30` >= by_stratum$`CrCl 30-60`),
  all(by_stratum$`CrCl 30-60`   >= by_stratum$`CrCl 60-90`),
  all(by_stratum$`CrCl 90-256`  <= by_stratum$`CrCl 60-90`)
)

# 1b. Magnitude of the renal effect, on the CENTRE of the regimen panel rather
#     than on any single arm: several arms saturate at 100% in the lowest
#     stratum, which compresses their individual gaps.
at_mic1 <- pta |> dplyr::filter(mic == 1)
gap <- median(at_mic1$pta[at_mic1$stratum == "CrCl 17.8-30"]) -
  median(at_mic1$pta[at_mic1$stratum == "CrCl 90-256"])
gap
stopifnot(gap > 20)

# 2. Dose linearity. The model is linear in dose and IIV sits on disposition
#    only, so doubling the dose is EXACTLY equivalent to doubling the MIC.
#    Because every arm draws the same cohort (common random numbers), this is
#    an exact equality, not an approximation -- and it is machine-stable
#    because it compares two arms of the same run.
lin <- pta |>
  dplyr::mutate(
    dose = ifelse(grepl("^1 g", treatment), 1000, 500),
    tau = as.integer(sub(".*q(\\d+)h", "\\1", treatment))
  )
lo <- lin |> dplyr::filter(dose == 500) |> dplyr::select(stratum, tau, mic, pta_lo = pta)
hi <- lin |> dplyr::filter(dose == 1000) |> dplyr::mutate(mic = mic / 2) |>
  dplyr::select(stratum, tau, mic, pta_hi = pta)
lin_chk <- dplyr::inner_join(lo, hi, by = c("stratum", "tau", "mic"))
stopifnot(nrow(lin_chk) > 20, identical(lin_chk$pta_lo, lin_chk$pta_hi))

# 3. Dosing frequency. At a fixed dose, shortening the interval raises the
#    trough, so PTA is non-decreasing as tau shrinks.
freq_chk <- lin |>
  dplyr::select(stratum, dose, tau, mic, pta) |>
  tidyr::pivot_wider(names_from = tau, values_from = pta, names_prefix = "tau")
stopifnot(all(freq_chk$tau6 >= freq_chk$tau8), all(freq_chk$tau8 >= freq_chk$tau12))
```

The strata reproduce the paper's two qualitative claims: PTA is markedly
higher in the lowest-CrCl stratum (Figure 4A) than in the others, and the
three strata above 30 mL/min sit much closer together than any of them sits to
the lowest (Figure 4B-D). The gradient across the upper three strata is more
visible here than in Figure 4 -- see Errata.

# Assumptions and deviations

## Modelling choices

* **The infusion rate `K0` is supplied by the event table, not by the model.**
  Eq. 3 carries `K0` as an explicit term; in `rxode2` a zero-order input is a
  property of the dose record (`dur = 1`), so `K0` does not appear in
  `model()`. This is a notational difference, not a structural one -- the
  closed-form check above confirms the two agree.
* **Clearance form.** Eqs. 3-6 are written in the clearance parameterisation
  (`CL`, `Q`, `Vc`, `Vp`). The model file uses the algebraically identical
  micro-constant form (`kel = cl/vc`, `k12 = q/vc`, `k21 = q/vp`) so that the
  ODE right-hand sides are in the package's usual shape; the closed-form check
  confirms the equivalence numerically.
* **IIV scale.** Table 2's variability column is headed `Iiv (CV%)`, so the
  tabulated numbers are apparent coefficients of variation of the log-normal
  distribution and are converted with `omega^2 = log(CV^2 + 1)`. Two features
  of the table confirm the reading: the residual-error row uses the same
  column and is separately labelled a proportional CV%, and the Table 2
  footnote quotes the CL entry as a percentage ("The IIV decreased from
  38.562% (base model) to 35.748% (final model)").
* **Omega structure.** The paper reports one variance per parameter and no
  covariances, so the omega matrix is encoded as diagonal.
* **No protein binding.** The paper's target-attainment discussion is framed
  in terms of *free* drug (`fT>MIC`) but no unbound fraction is reported
  anywhere, and no free-drug term appears in Eqs. 3-8. The model therefore
  predicts total plasma concentration, and every `fT>MIC` calculation in this
  vignette is on total concentration -- as the paper's own simulations must
  have been.
* **Cohort CrCl distribution.** Bai 2024 does not report the within-stratum
  CrCl distributions used for Figure 4. The cohort here is a log-normal
  matched to the Table 1 marginal mean and SD (104.59 and 60.95 mL/min),
  truncated to the observed 17.80-256.22 range and split at the paper's
  stratum boundaries. The upper stratum (90-256.22 mL/min) is wide, so the PTA
  in it is sensitive to this assumption; that is the most likely reason the
  gradient across strata B-D is more pronounced here than in Figure 4. The
  deterministic regimen check above does not depend on this assumption.
* **Extrapolation limit.** The paper states that "the predication model can
  only be used within the range (17.80-256.22 mL/min)". Every CrCl used in
  this vignette is inside that range.

## Errata and internal inconsistencies in the source

None of these change a parameter value; they are recorded so a later reader
does not mistake them for transcription errors here.

* **Table 1 median CrCl (99.34) is not the Eq. 8 centring constant (99.896).**
  The text below Eq. 8 calls 99.896 "the median value of CrCl", which
  disagrees with Table 1 by 0.56 mL/min. The model uses 99.896, because that
  is the value the fitted typical CL of 11.357 L/h is conditioned on. The
  difference is immaterial: at the exponent 0.473 it moves CL by 0.26%.
* **"First-order absorption".** Results, 'Population pharmacokinetics' opens
  "A two-compartment model with first-order absorption and elimination". There
  is no absorption in this study -- every dose was an intravenous infusion and
  Eq. 3 carries a zero-order `K0` input with no depot. The Abstract and
  Methods both describe the model correctly as "zero-order infusion and
  first-order elimination". The model has no depot compartment.
* **Eq. 4 initial condition.** Eq. 4 (the peripheral compartment) is printed
  with the initial condition `[Xc,0 = 0]`, repeating Eq. 3's label; it should
  read `Xp,0 = 0`. Both states start empty either way.
* **Dosing regimen typo.** Methods, 'Administration of the medications' lists
  "1 g every 6 hours (q8h)" -- the words and the abbreviation disagree. The
  Results and Discussion consistently treat 1 g q8h as the prescribed high
  regimen, and the simulation grid covers q6h, q8h and q12h regardless.
* **CrCl units in the Discussion.** The Discussion twice writes "CrCl >30
  mL/h" and "CrCl was less than 30 mL/h"; the units are mL/min everywhere
  else, including Table 1, Eq. 8 and the Results stratification.
* **Table 1 body weight.** Reported as 70.21 +/- 72.01 kg with a range of
  19.6-311.7 kg. The SD exceeds the mean and 311.7 kg is not plausible for
  this cohort, so the weight row should not be reused as a simulation
  distribution. Weight is not a covariate in the final model, so nothing in
  the model depends on it. (Weight does enter the Cockcroft-Gault CrCl the
  paper computed, but only the resulting CrCl is used here.)
* **Table 1 bloodstream infection.** "2 (0.04%)" -- 2 of 51 is 3.9%.
* **Table 1 stray footnote.** The line "Hepatic insufficiency: Child-Pugh
  B&C." appears inside the Results text immediately before Table 1 rather than
  in the table footnote block; it is a footnote to the "No. Hepatic
  insufficiency" row.
* **Supplemental figures.** Online supplemental figures 1 and 2 (ETA-versus-
  covariate scatter plots and a covariate correlation matrix) are cited but
  are diagnostics only; they carry no parameter values, so nothing in the
  model depends on them.
