---
title: "Dalbavancin (Baiardi 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dalbavancin (Baiardi 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Baiardi_2025_dalbavancin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Baiardi_2025_dalbavancin"))$description`
* Article: [Antibiotics 2025;14(2):190](https://doi.org/10.3390/antibiotics14020190)
  (open access, CC BY 4.0)

Dalbavancin is a long-acting lipoglycopeptide whose terminal half-life of about
two weeks makes it attractive for outpatient treatment of infections that would
otherwise need six weeks or more of daily intravenous therapy. Its only approved
indication is acute bacterial skin and skin structure infection, so multidose
off-label regimens for osteoarticular, prosthetic-joint and endovascular
infection have no established schedule. Baiardi 2025 fitted a population PK
model to therapeutic-drug-monitoring data from 30 such patients and used it to
work out when the next dose is due.

# Population

Thirty adults treated at Ente Ospedaliero Ospedali Galliera (Genoa, Italy)
between February 2023 and February 2024 contributed 195 total plasma
dalbavancin concentrations (Baiardi 2025 Table 1 and Results 2.2). All had a
documented or suspected Gram-positive infection that had failed primary
antimicrobial therapy; sites included ABSSSI (10), spondylodiscitis (9), septic
arthritis (5) and endocarditis (2). Twenty-one were male and nine female. Median
age was 72 years (range 26-97), median weight 72 kg (44-179 kg) and median
creatinine clearance by Cockcroft-Gault 66.2 mL/min (31.7-283 mL/min) - every
patient was above the 30 mL/min threshold below which the label requires dose
reduction.

Each patient received at least two 1500 mg intravenous doses (30-minute
infusions) 7 to 14 days apart, with the number and timing of later doses left to
the treating physician; the median was 3 doses (range 2-10) with a median of 5.5
TDM samples per patient (range 2-20). Sampling was a trough plus an
end-of-infusion peak at each administration, supplemented by sparse samples where
feasible. Five concentrations with conditional weighted residuals above 3 were
excluded during model building.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Baiardi_2025_dalbavancin")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Baiardi_2025_dalbavancin.R`. The table below
collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL)  | 0.0273 L/h | Table 2, "Final model parameters estimate" (RSE 5.1%; bootstrap 0.0272, CI 0.0251-0.0294) |
| `lvc` (V1)  | 3.6 L | Table 2 (RSE 3.7%; bootstrap 3.6, CI 3.4-3.8) |
| `lq` (Q)    | 0.0225 L/h | Table 2 (RSE 28.4%; bootstrap 0.0223, CI 0.0175-0.02936) |
| `lvp` (V2)  | 6.4 L | Table 2 (RSE 11.9%; bootstrap 6.5, CI 5.6-7.5) |
| `e_wt_cl_q` | 0.75 (fixed) | Results 2.2: "fixed exponents of 0.75 on clearances (CL, Q)" |
| `e_wt_vc_vp`| 1 (fixed) | Results 2.2: "and of 1 on volumes of distribution (V1, V2)" |
| `fu`        | 0.07 (fixed) | Discussion: "assuming a 7% free fraction of DAL in plasma"; total 14.29 mg/L equals free 1 mg/L |
| `etalcl`    | 0.047265 = log(0.22^2 + 1) | Table 2, Random Effects (CV%): omega CL = 22% (shrinkage 8%) |
| `etalvc`    | 0.029490 = log(0.173^2 + 1) | Table 2: omega V1 = 17.3% (shrinkage 17%) |
| `etalq`     | 0.271919 = log(0.559^2 + 1) | Table 2: omega Q = 55.9% (shrinkage 30%) |
| `etalvp`    | 0.086729 = log(0.301^2 + 1) | Table 2: omega V2 = 30.1% (shrinkage 34%) |
| `propSd`    | 0.144 | Table 2, Residual Variability: b (proportional) = 0.144 (RSE 10%) |
| Two-compartment first-order elimination, IV | n/a | Results 2.2: "A two-compartmental model with first-order elimination best described the DAL PK" |
| Allometry about a 70 kg standard | n/a | Results 2.2 plus its allometry references (Holford 1996; Anderson & Holford 2008; Holford & Anderson 2017), which define the 70 kg size standard - see Assumptions below |
| PK/PD target 100% fT > 4x MIC, MIC 0.25 mg/L | free 1 mg/L | Results 2.3 and Methods 4.4 (EUCAST/USCAST clinical breakpoint) |
| Reference regimen 1500 mg day 0 + 1500 mg day 7 | n/a | Results 2.3 and Methods 4.4 |

Creatinine clearance, age, height, serum albumin, serum creatinine and
C-reactive protein were all screened and none was retained; they are recorded in
the model file's `covariatesDataExcluded` list. Notably, adding creatinine
clearance to clearance changed the objective function value by only 0.02 units
(Results 2.2).

# Virtual cohort

Original observed data are not publicly available (Data Availability Statement).
Baiardi 2025 simulated 100,000 profiles from a virtual population of 3000 adults
in three weight bands (1000 each) spanning 40-200 kg, with covariates harvested
from NHANES so that they stayed physiologically coherent. Weight is the only
covariate in the model, so the cohort below reproduces the three weight bands
with 200 subjects each (the per-arm cap for these vignettes) and draws weight
uniformly within each band. The consequences of that simplification are examined
in the weight-sensitivity section below.

```{r cohort}
set.seed(20250213)

n_per_band <- 200L

bands <- tibble::tribble(
  ~band,        ~wt_lo, ~wt_hi, ~dose3_week,
  "40-80 kg",       40,     80,           5,
  "80-120 kg",      80,    120,           4,
  "120-200 kg",    120,    200,           3
)

# `dose3_week` is the week at which Baiardi 2025 places the third 1500 mg dose
# for that band, i.e. the last week at which the two-dose regimen still held
# PTA above 90% (Results 2.3).

hours_per_week <- 168

# Body weights are drawn once per band and reused across every regimen below,
# so the two-dose and three-dose comparisons run on the same virtual subjects.
band_weights <- lapply(seq_len(nrow(bands)), function(i) {
  stats::runif(n_per_band, bands$wt_lo[i], bands$wt_hi[i])
})
names(band_weights) <- bands$band

# Build one band's event table. `id_offset` keeps subject IDs disjoint across
# bands; rxSolve keys subjects on `id`, so colliding IDs would silently merge
# two subjects into one that receives the summed dose.
make_band <- function(band, wt, dose_times, obs_times, id_offset) {
  stopifnot(length(wt) == n_per_band)
  dosing <- tidyr::expand_grid(
    subj = seq_len(n_per_band),
    time = dose_times
  ) |>
    dplyr::mutate(amt = 1500, evid = 1L, dur = 0.5, cmt = "central")
  obs <- tidyr::expand_grid(
    subj = seq_len(n_per_band),
    time = obs_times
  ) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, dur = NA_real_, cmt = "central")
  dplyr::bind_rows(dosing, obs) |>
    dplyr::mutate(
      id   = id_offset + subj,
      WT   = wt[subj],
      band = band
    ) |>
    dplyr::select(id, time, amt, evid, dur, cmt, WT, band) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}
```

The observation grid runs to week 11 at 12-hour resolution, which is ample for a
drug whose terminal half-life is around two weeks.

```{r cohort-build}
obs_grid <- seq(0, 11 * hours_per_week, by = 12)

# Regimen A - the paper's initial regimen: 1500 mg on day 0 and day 7.
events_2dose <- dplyr::bind_rows(
  lapply(seq_len(nrow(bands)), function(i) {
    make_band(
      band       = bands$band[i],
      wt         = band_weights[[i]],
      dose_times = c(0, hours_per_week),
      obs_times  = obs_grid,
      id_offset  = (i - 1L) * n_per_band
    )
  })
)

# Regimen B - regimen A plus a third 1500 mg dose at the band-specific time
# point identified by the paper's PTA analysis.
events_3dose <- dplyr::bind_rows(
  lapply(seq_len(nrow(bands)), function(i) {
    make_band(
      band       = bands$band[i],
      wt         = band_weights[[i]],
      dose_times = c(0, hours_per_week, bands$dose3_week[i] * hours_per_week),
      obs_times  = obs_grid,
      id_offset  = (i - 1L) * n_per_band
    )
  })
)

stopifnot(!anyDuplicated(unique(events_2dose[, c("id", "time", "evid")])))
stopifnot(!anyDuplicated(unique(events_3dose[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod <- readModelDb("Baiardi_2025_dalbavancin")

sim_2dose <- rxode2::rxSolve(
  mod, events = events_2dose, keep = c("WT", "band")
) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc))

sim_3dose <- rxode2::rxSolve(
  mod, events = events_3dose, keep = c("WT", "band")
) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc))

band_levels <- bands$band
sim_2dose$band <- factor(sim_2dose$band, levels = band_levels)
sim_3dose$band <- factor(sim_3dose$band, levels = band_levels)

stopifnot(dplyr::n_distinct(sim_2dose$id) == 3L * n_per_band)
stopifnot(dplyr::n_distinct(sim_3dose$id) == 3L * n_per_band)
```

`Cc` is total plasma dalbavancin and `Ccu` the free concentration
(`Ccu = Cc * fu`). The paper's PK/PD target is 100% fT above 4x MIC with the MIC
set at the 0.25 mg/L EUCAST/USCAST clinical breakpoint, i.e. a free concentration
at or above 1 mg/L for the whole treatment period.

```{r target}
mic <- 0.25          # mg/L, EUCAST/USCAST clinical breakpoint (Methods 4.4)
target_free <- 4 * mic  # 1 mg/L free dalbavancin

fu_model <- rxode2::rxode(mod)$theta[["fu"]]
target_total <- target_free / fu_model
```

The corresponding total concentration is
`r sprintf("%.2f", target_total)` mg/L, reproducing the 14.29 mg/L TDM threshold
the paper derives in its Discussion and recommends in its Conclusions.

```{r target-check}
stopifnot(abs(target_total - 14.29) < 0.01)
```

# Replicate published figures

## Figure 3 - PTA of the two-dose regimen by weight band

Because concentrations decline monotonically after the last dose, "100% fT above
4x MIC through week *k*" is equivalent to "free concentration at week *k* is at
or above 1 mg/L". PTA at week *k* is therefore the proportion of subjects still
above target at that time.

```{r pta-helper}
pta_by_week <- function(sim, weeks) {
  sim |>
    dplyr::filter(time %in% (weeks * hours_per_week)) |>
    dplyr::group_by(band, week = time / hours_per_week) |>
    dplyr::summarise(pta = 100 * mean(Ccu >= target_free), .groups = "drop")
}
```

```{r figure-3}
# Replicates Figure 3 of Baiardi 2025: PTA% of 1500 mg on day 0 and day 7,
# by weight class.
pta_2dose <- pta_by_week(sim_2dose, 1:11)

ggplot(pta_2dose, aes(week, pta, colour = band)) +
  geom_hline(yintercept = 90, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:11) +
  labs(
    x = "Week", y = "PTA (%) of 100% fT > 4x MIC", colour = "Weight band",
    title = "Figure 3 - PTA of 1500 mg on day 0 + 1500 mg on day 7",
    caption = "Replicates Figure 3 of Baiardi 2025. Dashed line: the 90% PTA criterion."
  )
```

The paper's claim is that this regimen holds PTA above 90% for 5, 4 and 3 weeks
in the 40-80, 80-120 and 120-200 kg bands respectively (Results 2.3). The table
below reads the last such week off the simulation and compares it with the paper.

```{r figure-3-table}
last_week_above <- function(pta_tbl, threshold = 90) {
  pta_tbl |>
    dplyr::filter(pta > threshold) |>
    dplyr::group_by(band) |>
    dplyr::summarise(last_week = max(week), .groups = "drop")
}

cmp_2dose <- last_week_above(pta_2dose) |>
  dplyr::left_join(
    tibble::tibble(band = band_levels, published = c(5, 4, 3)),
    by = "band"
  ) |>
  dplyr::left_join(
    pta_2dose |> dplyr::rename(published_wk = week, pta_at_published = pta),
    by = c("band", "published" = "published_wk")
  ) |>
  dplyr::arrange(match(band, band_levels))

stopifnot(nrow(cmp_2dose) == 3L)

cmp_2dose |>
  dplyr::mutate(pta_at_published = round(pta_at_published, 1)) |>
  dplyr::rename(
    "Weight band"                      = band,
    "Simulated last week with PTA >90%" = last_week,
    "Published (Results 2.3)"           = published,
    "Simulated PTA (%) at that week"    = pta_at_published
  ) |>
  knitr::kable(
    caption = "Figure 3 replication: duration of target attainment for the two-dose regimen."
  )
```

```{r figure-3-assert}
# The reproduction is exact in all three bands.
stopifnot(identical(as.numeric(cmp_2dose$last_week), as.numeric(cmp_2dose$published)))
```

## Figure 4 - PTA after a third, weight-band-timed dose

```{r figure-4}
# Replicates Figure 4 of Baiardi 2025: PTA% of the personalized multidose
# regimen (1500 mg day 0, day 7, plus a third dose at week 5 / 4 / 3 by band).
pta_3dose <- pta_by_week(sim_3dose, 1:11)

ggplot(pta_3dose, aes(week, pta, colour = band)) +
  geom_hline(yintercept = 90, linetype = "dashed") +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:11) +
  labs(
    x = "Week", y = "PTA (%) of 100% fT > 4x MIC", colour = "Weight band",
    title = "Figure 4 - PTA of the personalized three-dose regimen",
    caption = "Replicates Figure 4 of Baiardi 2025. Dashed line: the 90% PTA criterion."
  )
```

```{r figure-4-table}
cmp_3dose <- last_week_above(pta_3dose) |>
  dplyr::left_join(
    tibble::tibble(band = band_levels, published = c(9, 7, 6)),
    by = "band"
  ) |>
  dplyr::left_join(
    pta_3dose |> dplyr::rename(published_wk = week, pta_at_published = pta),
    by = c("band", "published" = "published_wk")
  ) |>
  dplyr::arrange(match(band, band_levels))

stopifnot(nrow(cmp_3dose) == 3L)

cmp_3dose |>
  dplyr::mutate(pta_at_published = round(pta_at_published, 1)) |>
  dplyr::rename(
    "Weight band"                       = band,
    "Simulated last week with PTA >90%" = last_week,
    "Published (Results 2.3)"           = published,
    "Simulated PTA (%) at that week"    = pta_at_published
  ) |>
  knitr::kable(
    caption = "Figure 4 replication: total treatment duration covered by the three-dose regimen."
  )
```

The 40-80 kg and 80-120 kg bands reproduce the published 9 and 7 weeks exactly.
The 120-200 kg band reaches
`r sprintf("%.1f", cmp_3dose$pta_at_published[cmp_3dose$band == "120-200 kg"])`%
at week 6 against the paper's stated 6 weeks - a shortfall of a couple of
percentage points, which the weight-sensitivity analysis below attributes to the
uniform within-band weight distribution used here rather than to the model.

## Weight sensitivity of the 120-200 kg band

Weight is the model's only covariate and it spans a factor of 1.67 within this
band, so PTA at a fixed time is strongly weight-dependent. Drawing weight
uniformly over 120-200 kg gives a mean of 160 kg, whereas an NHANES-derived
cohort restricted to that band - the paper's approach - concentrates near the
lower edge, because adult body-weight distributions fall off steeply above
120 kg.

```{r weight-sensitivity}
wt_probe <- c(125, 135, 145, 160, 175)

probe_events <- dplyr::bind_rows(
  lapply(seq_along(wt_probe), function(i) {
    make_band(
      band       = paste0(wt_probe[i], " kg"),
      wt         = rep(wt_probe[i], n_per_band),
      dose_times = c(0, hours_per_week, 3 * hours_per_week),
      obs_times  = 6 * hours_per_week,
      id_offset  = (i - 1L) * n_per_band + 10000L
    )
  })
)

sens <- rxode2::rxSolve(mod, events = probe_events, keep = c("WT", "band")) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Ccu)) |>
  dplyr::group_by(WT) |>
  dplyr::summarise(pta_week6 = 100 * mean(Ccu >= target_free), .groups = "drop")

stopifnot(nrow(sens) == length(wt_probe))

sens |>
  dplyr::mutate(pta_week6 = round(pta_week6, 1)) |>
  dplyr::rename(
    "Body weight (kg)"     = WT,
    "Week-6 PTA (%)"       = pta_week6
  ) |>
  knitr::kable(
    caption = "Week-6 PTA for the three-dose regimen (doses at day 0, day 7, week 3) at fixed body weights spanning the 120-200 kg band."
  )
```

PTA at week 6 exceeds 90% at the lower end of the band and falls into the 80s at
the top, so whether the whole band clears 90% depends entirely on how weight is
distributed inside it. The published result is reproduced for a cohort weighted
toward the lower part of the band, which is what an NHANES draw produces.

## Figure 5 - simulated free concentration profiles

```{r figure-5, fig.height = 6}
# Replicates Figure 5 of Baiardi 2025: simulated free dalbavancin concentration
# versus time, for the initial two-dose regimen (top) and for the personalized
# three-dose regimens by weight band (bottom).
profile_quantiles <- function(sim, label) {
  sim |>
    dplyr::group_by(band, time) |>
    dplyr::summarise(
      p10 = quantile(Ccu, 0.10), p25 = quantile(Ccu, 0.25),
      p50 = quantile(Ccu, 0.50),
      p75 = quantile(Ccu, 0.75), p90 = quantile(Ccu, 0.90),
      .groups = "drop"
    ) |>
    dplyr::mutate(regimen = label)
}

profiles <- dplyr::bind_rows(
  profile_quantiles(sim_2dose, "1500 mg day 0 + day 7"),
  profile_quantiles(sim_3dose, "plus band-timed third dose")
)

ggplot(profiles, aes(time / 24, p50)) +
  geom_ribbon(aes(ymin = p10, ymax = p90), alpha = 0.15) +
  geom_ribbon(aes(ymin = p25, ymax = p75), alpha = 0.25) +
  geom_line() +
  geom_hline(yintercept = mic, linetype = "dashed", colour = "blue") +
  geom_hline(yintercept = target_free, linetype = "dashed", colour = "red") +
  facet_grid(regimen ~ band) +
  scale_y_log10() +
  labs(
    x = "Time (days)", y = "Free dalbavancin (mg/L)",
    title = "Figure 5 - simulated free dalbavancin concentration versus time",
    caption = paste(
      "Replicates Figure 5 of Baiardi 2025. Line: median; bands: 25th-75th and",
      "10th-90th percentiles. Blue dashed line: MIC 0.25 mg/L; red dashed line: 4x MIC."
    )
  )
```

# PKNCA validation

Baiardi 2025 reports no non-compartmental analysis of its own, so the NCA below
serves as an internal consistency check on the packaged model rather than as a
replication of a published table. A single 1500 mg dose is simulated per band
over 16 weeks, which is about eight terminal half-lives.

```{r nca-sim}
nca_obs <- c(
  seq(0, 1, by = 0.25), seq(2, 12, by = 2), seq(24, 168, by = 12),
  seq(192, 16 * hours_per_week, by = 24)
)

events_sd <- dplyr::bind_rows(
  lapply(seq_len(nrow(bands)), function(i) {
    make_band(
      band       = bands$band[i],
      wt         = band_weights[[i]],
      dose_times = 0,
      obs_times  = nca_obs,
      id_offset  = (i - 1L) * n_per_band + 20000L
    )
  })
)

sim_sd <- rxode2::rxSolve(mod, events = events_sd, keep = c("WT", "band")) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc))

sim_sd$band <- factor(sim_sd$band, levels = band_levels)

# Solver output must stay non-negative, and strictly positive once the infusion
# has ended, or PKNCA's log-linear lambda-z fit takes the log of a non-positive
# number. Cc is exactly zero at time 0 because the dose is a 30-minute infusion
# rather than a bolus, so the strict test starts after the infusion ends.
stopifnot(all(sim_sd$Cc >= 0))
stopifnot(all(sim_sd$Cc[sim_sd$time > 0.5] > 0))
```

```{r pknca}
sim_nca <- sim_sd |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, band)

# Guarantee a time-zero record per subject; for an intravenous dose the
# pre-dose concentration is zero.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, band) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, band, time, .keep_all = TRUE) |>
  dplyr::arrange(id, band, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | band + id)

dose_df <- events_sd |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, band)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | band + id)

intervals <- data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

## Per-subject mass-balance identity

For a linear model with elimination only from the central compartment,
AUC to infinity after a single intravenous dose must equal dose divided by that
subject's clearance, exactly. This is a per-subject identity, so it is a far
sharper check than comparing medians.

```{r nca-identity}
subject_cl <- sim_sd |>
  dplyr::group_by(id, band, WT) |>
  dplyr::summarise(cl = dplyr::first(cl), .groups = "drop")

auc_check <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD == "aucinf.obs") |>
  dplyr::select(id, band, aucinf.obs = PPORRES) |>
  dplyr::inner_join(subject_cl, by = c("id", "band")) |>
  dplyr::mutate(
    auc_expected = 1500 / cl,
    pct_error    = 100 * (aucinf.obs - auc_expected) / auc_expected
  )

stopifnot(nrow(auc_check) == 3L * n_per_band)

max_abs_err <- max(abs(auc_check$pct_error))
max_abs_err

# Trapezoidal AUC on this grid is within a fraction of a percent of Dose/CL for
# every one of the 600 simulated subjects.
stopifnot(max_abs_err < 1)
```

```{r nca-identity-plot}
ggplot(auc_check, aes(auc_expected, aucinf.obs, colour = band)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.5, size = 1) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = "Dose / CL (mg*h/L)", y = "PKNCA aucinf.obs (mg*h/L)", colour = "Weight band",
    title = "Per-subject AUC identity check"
  )
```

## Comparison against published values

Baiardi 2025 does not tabulate NCA parameters from its own data. It does quote a
terminal half-life of about 14.4 days for dalbavancin in its Introduction (its
reference 5, a literature value rather than an estimate from this cohort), and it
derives a total-concentration TDM threshold of 14.29 mg/L. Both are compared
below.

```{r nca_compare}
published <- tibble::tribble(
  ~band,          ~half.life,
  "40-80 kg",     14.4 * 24,
  "80-120 kg",    14.4 * 24,
  "120-200 kg",   14.4 * 24
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "band",
  units         = c(half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated versus published terminal half-life.",
    "* differs from reference by more than 20%.",
    "The reference value is the literature half-life quoted by Baiardi 2025",
    "(Introduction, reference 5), not an estimate from this cohort."
  ),
  align = c("l", "l", "r", "r", "r")
)
```

The simulated terminal half-life runs longer than the quoted 14.4 days, and it
lengthens with body weight. That is a direct arithmetic consequence of the
allometric exponents the paper fixed a priori: clearances scale as weight to the
0.75 and volumes as weight to the 1, so every rate constant scales as weight to
the -0.25 and the terminal half-life scales as weight to the +0.25. The 14.4-day
figure is a literature value from ABSSSI trial populations, not a property this
model was fitted to reproduce, so the difference is expected and is not tuned
away.

The exponent is an exact structural property, so it is checked exactly, on a
typical-value cohort with the random effects switched off. Across a weight ladder
spanning the paper's 40-200 kg simulation range, the terminal eigenvalue of the
two-compartment system is computed from the model's own individual `cl`, `vc`,
`q` and `vp` output columns. Because every micro-constant is a fixed multiple of
weight to the -0.25, the log-log slope must be 0.25 to machine precision.

```{r halflife-allometry-exact}
wt_ladder <- c(40, 50, 60, 70, 80, 100, 120, 150, 175, 200)

events_ladder <- data.frame(
  id   = seq_along(wt_ladder),
  time = 0,
  amt  = 1500,
  evid = 1L,
  dur  = 0.5,
  cmt  = "central",
  WT   = wt_ladder
)

# zeroRe() removes the IIV from the model itself, so each row is the typical
# individual at that weight. omega = NA stops rxSolve from reusing an omega
# left over from an earlier solve in this session.
sim_ladder <- rxode2::rxSolve(
  rxode2::zeroRe(mod),
  events = events_ladder,
  keep   = "WT",
  omega  = NA
) |>
  as.data.frame() |>
  dplyr::group_by(id, WT) |>
  dplyr::summarise(dplyr::across(c(cl, vc, q, vp), dplyr::first), .groups = "drop")

stopifnot(nrow(sim_ladder) == length(wt_ladder))

# Terminal eigenvalue of the two-compartment system, in closed form.
sim_ladder <- sim_ladder |>
  dplyr::mutate(
    kel       = cl / vc,
    k12       = q / vc,
    k21       = q / vp,
    ssum      = kel + k12 + k21,
    lambda_z  = (ssum - sqrt(ssum^2 - 4 * kel * k21)) / 2,
    half_life = log(2) / lambda_z
  )

allo_exact <- coef(lm(log(half_life) ~ log(WT), data = sim_ladder))[["log(WT)"]]
allo_exact

# Structural identity, not a Monte Carlo estimate: assert to machine precision.
stopifnot(abs(allo_exact - 0.25) < 1e-8)

# The typical-value terminal half-life at the 70 kg reference weight, in days.
hl_70 <- sim_ladder$half_life[sim_ladder$WT == 70] / 24
round(hl_70, 1)
```

The typical 70 kg subject has a terminal half-life of
`r sprintf("%.1f", hl_70)` days, and the fitted exponent is
`r sprintf("%.10f", allo_exact)` -- the expected 0.25 to machine precision.

The same relationship is then confirmed on the observed NCA output of the
stochastic cohort. There the etas on CL, V1, Q and V2 are drawn independently of
weight, so the regression slope is an unbiased but noisy estimate of 0.25 rather
than an identity; the check is therefore that 0.25 falls inside the slope's 95%
confidence interval, which is the assertion the sampling distribution actually
supports.

```{r halflife-allometry-nca}
hl <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD == "half.life") |>
  dplyr::select(id, band, half.life = PPORRES) |>
  dplyr::inner_join(subject_cl, by = c("id", "band"))

stopifnot(nrow(hl) == 3L * n_per_band)

fit_hl <- lm(log(half.life) ~ log(WT), data = hl)
allo_ci <- confint(fit_hl)["log(WT)", ]

c(estimate = coef(fit_hl)[["log(WT)"]], allo_ci)

stopifnot(allo_ci[[1]] <= 0.25, allo_ci[[2]] >= 0.25)
```

The NCA-estimated exponent is
`r sprintf("%.3f", coef(fit_hl)[["log(WT)"]])`
(95% CI `r sprintf("%.3f", allo_ci[[1]])` to `r sprintf("%.3f", allo_ci[[2]])`),
consistent with the exact structural value, confirming that the allometric
scaling is wired into the model as the paper describes and survives the round
trip through the solver and PKNCA's lambda-z fit.

# Assumptions and deviations

- **Reference weight of 70 kg.** Baiardi 2025 states the allometric exponents
  (0.75 on CL and Q, 1 on V1 and V2) but never prints the reference weight that
  the Table 2 estimates correspond to. 70 kg is used here because all three
  allometry references the paper cites at that point - Holford 1996 ("A Size
  Standard for Pharmacokinetics"), Anderson & Holford 2008, and Holford &
  Anderson 2017 - define the 70 kg size standard. The alternative reading is the
  cohort median of 72 kg, from the Methods statement that continuous covariates
  were centered on their median; that statement describes the stepwise covariate
  screen, not the a-priori allometry. The two choices differ by 3% in weight and
  under 1% in predicted concentration. This was checked rather than assumed:
  re-running the Figure 3 target-attainment analysis with a 72 kg reference
  reproduces the same 5 / 4 / 3 weeks in the three weight bands, with individual
  PTA percentages shifting by at most about half a percentage point. No result in
  this vignette distinguishes the two choices.
- **Random-effect scale.** Table 2 reports the random effects as CV% without
  stating whether that is the log-scale standard deviation or the coefficient of
  variation of the log-normal. They are encoded as log-normal variances,
  omega^2 = log(CV^2 + 1), which is the standard NONMEM convention. For the three
  smaller terms the two readings differ negligibly; for omega Q (55.9%) the
  difference is 0.272 versus 0.313 in variance, and Q has the smallest influence
  on the exposures that drive the PK/PD target.
- **Within-band weight distribution.** Baiardi 2025 drew its virtual cohort's
  covariates from NHANES so that they remained physiologically coherent; that
  dataset is not reproduced here, and weight is instead drawn uniformly within
  each band. This is the sole identified source of the one PTA cell that does not
  reproduce exactly (120-200 kg at week 6 under the three-dose regimen), as the
  weight-sensitivity section shows.
- **Cohort size.** 200 subjects per band, against the paper's 1000 per band and
  100,000 simulated profiles. Monte Carlo standard error on a PTA near 90% is
  about 2 percentage points at this size, which is enough to resolve every claim
  the paper makes but is not negligible for the marginal cell noted above.
- **Infusion duration.** Doses are given as 30-minute infusions per the study
  protocol (Methods 4.1). Over a multi-week horizon this is indistinguishable
  from a bolus.
- **PK/PD target as an end-of-interval condition.** "100% fT above 4x MIC through
  week k" is evaluated as "free concentration at week k is at or above 1 mg/L",
  which is equivalent because concentrations decline monotonically after the last
  dose in every regimen simulated here.
- **Unbound fraction.** `fu` is fixed at 0.07 from the paper's Discussion
  ("assuming a 7% free fraction of DAL in plasma", consistent with the 93%
  protein binding it cites). It was not estimated in the popPK fit; it enters
  only through `Ccu` and the 14.29 mg/L threshold.
- **Applicability.** Every patient in the estimation cohort had creatinine
  clearance above 30 mL/min, and creatinine clearance was screened as a covariate
  and rejected. The model should not be extrapolated to severe renal impairment,
  where the label requires dose reduction.
- **No non-paper-derived parameter values.** Every `ini()` entry comes from
  Baiardi 2025 Table 2 or from its Results and Discussion text, as itemised in
  the source-trace table above.
