---
title: "Dabrafenib, hydroxy-dabrafenib and trametinib (Balakirouchenane 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dabrafenib, hydroxy-dabrafenib and trametinib (Balakirouchenane 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Models and source

```{r models}
mod_dab <- rxode2::rxode(readModelDb("Balakirouchenane_2020_dabrafenib"))
mod_tra <- rxode2::rxode(readModelDb("Balakirouchenane_2020_trametinib"))
```

* Citation: `r mod_dab$reference`
* Article: <https://doi.org/10.3390/cancers12040931>

Balakirouchenane and colleagues fit two population PK models to routine-care
samples from French patients treated with the BRAF inhibitor dabrafenib, alone
or combined with the MEK inhibitor trametinib, and then related the model-based
exposures to dose-limiting toxicity and survival. The two PK models were fit
separately and are shipped as two files that share this article:

* `Balakirouchenane_2020_dabrafenib` -- `r mod_dab$description`
* `Balakirouchenane_2020_trametinib` -- `r mod_tra$description`

The exposure-response part of the paper (Fisher / Wilcoxon tests for
dose-limiting toxicity, Cox models for overall and progression-free survival)
is a statistical analysis of per-patient AUCs and has no structural model to
encode.

# Population

Seventy-three adults with BRAF V600-mutated metastatic solid tumours contributed
424 dabrafenib / hydroxy-dabrafenib records; the 60 who also took trametinib
contributed 318 trametinib concentrations (Table 1). Most had metastatic
melanoma (89% in the dabrafenib set, 87% in the trametinib set); the rest had
anaplastic thyroid or non-small-cell lung carcinoma. In the dabrafenib set the
median (range) age was 61.2 (20-90) years and body weight 73.0 (51.7-166.0) kg,
and 41% were women; 78% took a proton-pump inhibitor. Patients started
dabrafenib at 150 mg twice daily and trametinib at 2 mg once daily, with reduced
starting doses allowed. All samples were drawn at steady state during routine
visits at three Paris hospitals between July 2015 and June 2017.

```{r population}
str(mod_dab$population[c("n_subjects", "age_range", "sex_female_pct", "dose_range")])
```

# Source trace

| Element | Value | Source |
|---|---|---|
| Dabrafenib / OHD structure: 2-cmt parent, lag + first-order absorption, complete conversion to 2-cmt metabolite | -- | Results 2.2.1; Figure 1; Methods 4.4.1 |
| `lka` (fixed) | 1.8 1/h | Table 2; Methods 4.4.1 |
| `ltlag` | 0.499 h | Table 2 |
| `lcl`, `lvc`, `lq`, `lvp` | 19.3 L/h, 39.1 L, 3.40 L/h, 18.7 L | Table 2 (CL/F, V2/F, Q/F, V4/F) |
| `lcl_ohd`, `lvc_ohd`, `lq_ohd`, `lvp_ohd` | 23.2 L/h, 5.11 L, 7.21 L/h, 27.1 L | Table 2 (CLm/F, V3/F, Qm/F, V5/F) |
| `e_age_cl`, `e_age_cl_ohd` | -0.536, -0.589 | Table 2 (unsigned); sign from Results 2.2.1 and Figure 3 |
| `e_sexf_cl` | 0.832 | Table 2 |
| Covariate equations, MAGE = 61.2 y, Sex 1 = woman | -- | Table 2 footnote |
| `etalcl`, `etalvc`, `etalcl_ohd`, `etalvc_ohd` | 16.0, 50.8, 24.0, 47.5 %CV | Table 2 |
| `etaiov_cl_*` | 17.4 %CV | Table 2 'IOV' |
| `propSd`, `propSd_ohd` | 48.7%, 53.1% | Table 2 (correlation 87.0% not encoded) |
| Molecular weights 519.56 / 535.56 g/mol | -- | Computed from the formulas; analysis in molar units per Methods 4.4 |
| Trametinib structure: 2-cmt, lag + first-order absorption | -- | Results 2.2.2 |
| `lka`, `ltlag`, `lcl`, `lvc`, `lq`, `lvp` | 0.913 1/h, 0.709 h, 5.83 L/h, 61.9 L, 64.9 L/h, 417.0 L | Table 3 |
| `etalcl`, `etalq` | 29.6, 80.2 %CV | Table 3 |
| `addSd` | 4.14 ng/mL | Table 3 |

# Typical-value check against Figure 3

Figure 3 of the paper simulates 150 mg dabrafenib twice daily in men and women
aged 20 and 90 and reports the median composite AUC (dabrafenib +
hydroxy-dabrafenib over one dosing interval). With IIV switched off the model
should reproduce those medians, which makes this the most direct check of the
age and sex covariate equations, including the sign of the age coefficients
and the molar conversion from dabrafenib to hydroxy-dabrafenib.

```{r typical-fig3}
groups <- tibble::tribble(
  ~group,         ~AGE, ~SEXF, ~published,
  "Men, 20 y",      20,     0,      10447,
  "Men, 90 y",      90,     0,      19542,
  "Women, 20 y",    20,     1,      11632,
  "Women, 90 y",    90,     1,      21707
)

tau <- 12
ss_start <- 29 * 24 + 12 # start of the 60th dosing interval

make_dab_events <- function(n_per_group, groups) {
  base <- rxode2::et(amt = 150, ii = tau, addl = 59, cmt = "depot") |>
    # dense over the lag time and absorption peak, where a coarse grid biases
    # the trapezoidal AUC
    rxode2::et(ss_start + c(seq(0, 3, by = 0.05), seq(3.25, tau, by = 0.25))) |>
    as.data.frame()
  out <- lapply(seq_len(nrow(groups)), function(i) {
    ids <- (i - 1) * n_per_group + seq_len(n_per_group)
    do.call(rbind, lapply(ids, function(j) {
      d <- base
      d$id <- j
      d
    })) |>
      dplyr::mutate(group = groups$group[i], AGE = groups$AGE[i], SEXF = groups$SEXF[i])
  })
  dplyr::bind_rows(out) |>
    dplyr::mutate(
      OCC = 1L,
      # Two declared endpoints, neither an ODE state: key observation rows by dvid
      dvid = ifelse(evid == 0, 1L, NA_integer_),
      cmt = ifelse(evid == 0, NA_character_, cmt)
    )
}

trap <- function(t, y) sum(diff(t) * (head(y, -1) + tail(y, -1)) / 2)

ev_typ <- make_dab_events(1, groups)
sim_typ <- as.data.frame(rxode2::rxSolve(mod_dab, ev_typ, omega = NA, sigma = NA,
                                         keep = "group"))

typ <- sim_typ |>
  dplyr::group_by(group) |>
  dplyr::summarise(auc_dab = trap(time, Cc), auc_ohd = trap(time, Cc_ohd), .groups = "drop") |>
  dplyr::mutate(composite = auc_dab + auc_ohd) |>
  dplyr::left_join(groups, by = "group") |>
  dplyr::mutate(pct_diff = 100 * (composite - published) / published)

typ |>
  dplyr::select(group, auc_dab, auc_ohd, composite, published, pct_diff) |>
  dplyr::mutate(dplyr::across(c(auc_dab, auc_ohd, composite), round)) |>
  dplyr::rename(
    "Group" = group, "AUCtau DAB (ng.h/mL)" = auc_dab,
    "AUCtau OHD (ng.h/mL)" = auc_ohd, "Composite (ng.h/mL)" = composite,
    "Figure 3 median (ng.h/mL)" = published, "% difference" = pct_diff
  ) |>
  knitr::kable(digits = 2, caption = "Typical-value composite AUC vs Figure 3 medians.")

# Deterministic: a sign error on the age coefficients roughly halves or
# doubles these values, and dropping the molar conversion shifts them by ~1.5%.
stopifnot(all(abs(typ$pct_diff) < 1))
```

# Stochastic steady-state simulation (dabrafenib / hydroxy-dabrafenib)

Two hundred virtual patients per Figure 3 group receive 150 mg twice daily for
30 days, one occasion each. The composite AUC distribution per group is the
stochastic analogue of Figure 3.

```{r sim-dab}
rxode2::rxSetSeed(20200409)
n_per_group <- 200
ev_dab <- make_dab_events(n_per_group, groups)
sim_dab <- as.data.frame(rxode2::rxSolve(mod_dab, ev_dab, keep = "group",
                                         returnType = "data.frame"))
```

```{r plot-dab, fig.cap = "Steady-state profiles over one 12-h interval: median and 5th-95th percentiles of the individual predictions (no residual error), by Figure 3 group. The shape corresponds to the pcVPC of Figure 2."}
sim_dab |>
  dplyr::select(id, time, group, Cc, Cc_ohd) |>
  tidyr::pivot_longer(c(Cc, Cc_ohd), names_to = "analyte", values_to = "conc") |>
  dplyr::mutate(analyte = ifelse(analyte == "Cc", "Dabrafenib", "Hydroxy-dabrafenib"),
                tad = time - ss_start) |>
  dplyr::group_by(group, analyte, tad) |>
  dplyr::summarise(p05 = quantile(conc, 0.05), p50 = median(conc),
                   p95 = quantile(conc, 0.95), .groups = "drop") |>
  ggplot(aes(tad, p50, colour = group, fill = group)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.15, colour = NA) +
  geom_line() +
  facet_wrap(~analyte) +
  labs(x = "Time after dose (h)", y = "Concentration (ng/mL)", colour = NULL, fill = NULL) +
  theme_bw()
```

## PKNCA over the steady-state interval

```{r nca-dab}
nca_in <- sim_dab |>
  dplyr::select(id, time, group, Cc, Cc_ohd) |>
  tidyr::pivot_longer(c(Cc, Cc_ohd), names_to = "analyte", values_to = "conc") |>
  dplyr::filter(!is.na(conc)) |>
  dplyr::rename(treatment = group)

dose_dab <- ev_dab |>
  dplyr::filter(evid == 1) |>
  dplyr::transmute(id, time, amt, treatment = group)

conc_obj <- PKNCA::PKNCAconc(nca_in, conc ~ time | treatment + id / analyte)
dose_obj <- PKNCA::PKNCAdose(dose_dab, amt ~ time | treatment + id)
intervals <- data.frame(start = ss_start, end = ss_start + tau,
                        auclast = TRUE, cmax = TRUE, cmin = TRUE, tmax = TRUE)
nca_dab <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_dab) |>
  dplyr::select(treatment, id, analyte, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = c(analyte, PPTESTCD), values_from = PPORRES)
stopifnot(nrow(nca_wide) == 4 * n_per_group)
```

### Internal identity: steady-state AUCtau equals Dose / CL

At steady state the dabrafenib AUC over a dosing interval is exactly
`Dose / CL`, and, because conversion to the metabolite is complete, the
hydroxy-dabrafenib AUC is `Dose / CLm` scaled by the molecular-weight ratio.
Both sides use each subject's own drawn parameters, so the difference is pure
trapezoidal error and a tight bound is appropriate.

```{r identity-dab}
indiv <- sim_dab |>
  dplyr::distinct(id, cl, cl_ohd)
ident <- nca_wide |>
  dplyr::left_join(indiv, by = "id") |>
  dplyr::mutate(
    dab_pct = 100 * (Cc_auclast - 150 / cl * 1000) / (150 / cl * 1000),
    ohd_ref = 150 / cl_ohd * (535.56 / 519.56) * 1000,
    ohd_pct = 100 * (Cc_ohd_auclast - ohd_ref) / ohd_ref
  )
knitr::kable(tibble::tibble(
  Analyte = c("Dabrafenib", "Hydroxy-dabrafenib"),
  "Max absolute % difference" = c(max(abs(ident$dab_pct)), max(abs(ident$ohd_pct)))
), digits = 3, caption = "PKNCA AUCtau vs the analytic steady-state value.")
stopifnot(max(abs(ident$dab_pct)) < 1, max(abs(ident$ohd_pct)) < 1)
```

### Comparison against Figure 3

```{r compare-fig3}
sim_summary <- nca_wide |>
  dplyr::mutate(composite = Cc_auclast + Cc_ohd_auclast) |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(auclast = median(composite), .groups = "drop")

published <- groups |>
  dplyr::transmute(treatment = group, auclast = published)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_summary,
  reference = published,
  by = "treatment",
  params = "auclast",
  units = c(auclast = "ng.h/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = paste(
  "Median composite AUCtau (dabrafenib + hydroxy-dabrafenib) of 200 simulated",
  "patients per group vs the Figure 3 medians. * differs by >20%."
))

chk <- sim_summary |>
  dplyr::left_join(published, by = "treatment", suffix = c("_sim", "_pub")) |>
  dplyr::mutate(pct = 100 * (auclast_sim - auclast_pub) / auclast_pub)
# Median of 200 log-normal draws with ~20-30% CV: Monte-Carlo error ~2%.
stopifnot(all(abs(chk$pct) < 7))
```

The paper's Section 2.4 also reports the median model-estimated AUC over the
first three months in the 30 melanoma patients of the survival analysis:
7722 ng.h/mL for dabrafenib and 6082 ng.h/mL for hydroxy-dabrafenib, a
hydroxy-dabrafenib / dabrafenib ratio near the 0.80 the Discussion quotes.
Those are empirical Bayes estimates in a cohort that included reduced doses,
so they are context rather than a gate; the simulated 150 mg values at the
cohort's median age are of the same order.

```{r median-age}
ev_med <- make_dab_events(1, tibble::tibble(group = "Man, 61.2 y", AGE = 61.2, SEXF = 0))
sim_med <- as.data.frame(rxode2::rxSolve(mod_dab, ev_med, omega = NA, sigma = NA))
c(auc_dab = trap(sim_med$time, sim_med$Cc), auc_ohd = trap(sim_med$time, sim_med$Cc_ohd))
```

# Trametinib

```{r sim-tra}
rxode2::rxSetSeed(20200410)
n_tra <- 200
# 60 days: with IIV of 80% on Q/F and a 417 L peripheral volume, subjects with
# a low Q/F take several weeks to approach steady state.
ss_tra <- 59 * 24
ev_tra <- rxode2::et(amt = 2, ii = 24, addl = 59, cmt = "depot") |>
  rxode2::et(ss_tra + c(seq(0, 4, by = 0.05), seq(4.25, 24, by = 0.25)), cmt = "central") |>
  rxode2::et(id = seq_len(n_tra)) |>
  as.data.frame() |>
  dplyr::mutate(treatment = "2 mg once daily")
sim_tra <- as.data.frame(rxode2::rxSolve(mod_tra, ev_tra, keep = "treatment",
                                         returnType = "data.frame"))
```

```{r plot-tra, fig.cap = "Trametinib steady-state profile over one 24-h interval: median and 5th-95th percentiles of the individual predictions (compare Figure 4)."}
sim_tra |>
  dplyr::mutate(tad = time - ss_tra) |>
  dplyr::group_by(tad) |>
  dplyr::summarise(p05 = quantile(Cc, 0.05), p50 = median(Cc), p95 = quantile(Cc, 0.95)) |>
  ggplot(aes(tad, p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.2) +
  geom_line() +
  labs(x = "Time after dose (h)", y = "Trametinib (ng/mL)") +
  theme_bw()
```

```{r nca-tra}
conc_tra <- PKNCA::PKNCAconc(
  sim_tra |> dplyr::filter(!is.na(Cc)) |> dplyr::select(id, time, Cc, treatment),
  Cc ~ time | treatment + id
)
dose_tra <- PKNCA::PKNCAdose(
  ev_tra |> dplyr::filter(evid == 1) |> dplyr::select(id, time, amt, treatment),
  amt ~ time | treatment + id
)
int_tra <- data.frame(start = ss_tra, end = ss_tra + 24,
                      auclast = TRUE, cmax = TRUE, cmin = TRUE, tmax = TRUE)
nca_tra <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_tra, dose_tra, intervals = int_tra))
tra_wide <- as.data.frame(nca_tra) |>
  tidyr::pivot_wider(id_cols = c(treatment, id), names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::left_join(sim_tra |> dplyr::distinct(id, cl), by = "id") |>
  dplyr::mutate(pct = 100 * (auclast - 2 / cl * 1000) / (2 / cl * 1000))
# Dose / CL holds only at true steady state; the residual gap is the slow
# peripheral approach in low-Q/F subjects (a per-subject physical effect), so
# gate on the centre and a robust quantile rather than on the maximum.
stopifnot(
  nrow(tra_wide) == n_tra,
  abs(median(tra_wide$pct)) < 0.5,
  quantile(abs(tra_wide$pct), 0.9) < 1
)

tra_summary <- tra_wide |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(auclast = median(auclast), cmax = median(cmax),
                   cmin = median(cmin), .groups = "drop")
tra_pub <- tibble::tibble(treatment = "2 mg once daily", auclast = 292)

knitr::kable(nlmixr2lib::ncaComparisonTable(
  simulated = tra_summary, reference = tra_pub, by = "treatment",
  params = "auclast", units = c(auclast = "ng.h/mL"), tolerance_pct = 20
), caption = paste(
  "Median simulated trametinib AUCtau at 2 mg once daily vs the median",
  "first-three-month AUC of the 30-patient survival cohort (Section 2.4).",
  "* differs by >20%."
))
```

The simulated median AUCtau at 2 mg once daily (about 340 ng.h/mL, i.e.
Dose / CL = 2 mg / 5.83 L/h) is about 16% above the 292 ng.h/mL reported for
the survival cohort. That cohort included patients on reduced trametinib doses
(1 mg once daily for three patients at start, and dose reductions in 23% of
the melanoma cohort), which pull the empirical median down, so a modest
positive difference is the expected direction. Its median trough of about
`r round(tra_summary$cmin, 1)` ng/mL sits near the 10.6 ng/mL efficacy
threshold cited in the paper's Introduction.

# Assumptions and deviations

* **Sign of the age coefficients.** Table 2 prints the age effects on
  dabrafenib CL/F (0.536) and hydroxy-dabrafenib CLm/F (0.589) without a sign,
  and the footnote writes them as `1 + theta * (AGE - 61.2) / 61.2`. Taken
  literally this makes clearance rise with age, which contradicts the Results
  ("CL/F is reduced ... by 55% when comparing 20-year-old to 90-year-old
  patients"; "a similar decrease (51%) ... in CLm/F"), the Figure 3 AUCs, and
  the Discussion's advice to reduce the starting dose in the elderly. The
  coefficients are encoded as negative; with that sign the typical-value
  composite AUCs reproduce all four Figure 3 medians to within 0.5%.
* **Molar units.** The dabrafenib / hydroxy-dabrafenib model was fit in molar
  units. The shipped model takes doses in mg of dabrafenib, carries the
  metabolite states in dabrafenib-equivalent mg, and converts the
  hydroxy-dabrafenib concentration to ng/mL with the molecular-weight ratio
  535.56 / 519.56. The molecular weights are computed from the chemical
  formulas and are not stated in the paper.
* **Correlated residual error not encoded.** The two proportional errors were
  estimated with a correlation of 87% (NONMEM L2 item). nlmixr2 does not
  support correlated residual errors, so the shipped model treats them as
  independent; typical-value and individual predictions are unaffected, but
  simulated observations of the two analytes will be less correlated than in
  the original fit.
* **Occasions for the IOV on dabrafenib CL/F.** The paper estimates an IOV of
  17.4% without defining the occasions. Six occasion slots (`OCC` = 1-6) are
  provided, repeating the estimated variance; an `OCC` value outside that
  range gets no IOV. The simulations above use a single occasion.
* **Variability scale.** IIV and IOV are reported as CV%; they are converted
  with `omega^2 = log(1 + CV^2)`, the convention the same group uses in its
  later trametinib model. The alternative `omega = CV` reading would give
  variances up to 12% larger for the dabrafenib / hydroxy-dabrafenib terms
  and 29% larger for the IIV on trametinib Q/F (80.2%), the only term where
  the choice is material.
* **Trametinib residual error** is taken as an additive standard deviation of
  4.14 ng/mL (Table 3 'RUV (ng/mL)').
* **Virtual cohort.** The Figure 3 groups fix age and sex, as the paper does;
  no other covariate enters either model.
* **Table 1 typo.** The trametinib cohort's BSA is printed as '3.1 (2.5-3.0)',
  a median outside its range; BSA was not retained so this does not affect
  the model.
