---
title: "Busulfan (Balis 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Busulfan (Balis 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Balis_2025_busulfan"))$reference`
* Description: `r rxode2::rxode(readModelDb("Balis_2025_busulfan"))$description`
* Article: <https://doi.org/10.1007/s00280-025-04757-w>
* Supplement (Supplemental Tables 1-8 and Supplemental Figures 1-8): <https://doi.org/10.1007/s00280-025-04757-w> (Electronic Supplementary Material, files `280_2025_4757_MOESM1_ESM.pdf` and `280_2025_4757_MOESM2_ESM.pdf`)

# Population

Balis 2025 is a retrospective single-institution analysis of busulfan therapeutic
drug monitoring (TDM) data collected at the Children's Hospital of Philadelphia
between February 2007 and May 2023. Three hundred twenty-eight infants and
children with malignant (n = 172) and non-malignant (n = 156) disease received a
busulfan-containing hematopoietic stem cell transplant conditioning regimen and
had plasma busulfan monitored after their first dose. The median age was
5.9 years (range 0.11 to 28.1), median body weight 20.5 kg (range 3.4 to 126) and
median body surface area 0.81 m^2 (range 0.21 to 2.42); 195 patients were male
and 133 female (Balis 2025 Table 2). Twenty-seven percent of the population was
under 2 years of age.

Busulfan was given on one of two schedules: 3.2 mg/kg/dose infused over 3 h once
daily for 4 doses (n = 154), or 0.8 mg/kg/dose infused over 2 h every 6 h for 16
doses (n = 174); both deliver 12.8 mg/kg over the course. Thirty-eight children
who were > 10 kg and < 48 months of age on the daily schedule received 4 mg/kg,
and 44 patients on the q6h schedule received 1 mg/kg. Doses were adjusted from
day 2 onward to target an AUCinf between 14.8 and 24.6 mg*h/L (3600 to
6000 uM*min). Six or seven plasma samples were drawn after the first infusion
(Supplemental Table 2). Kidney and liver function laboratory values were within
the reference interval for essentially all subjects and were therefore not tested
as covariates.

The model was fitted in Phoenix NLME v8.3 using FOCE-ELS. The parameters packaged
here are the final covariate model of Supplemental Table 6, fitted to the
first-dose (day 1) data (2149 concentrations).

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Balis_2025_busulfan")()$population`).

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| One-compartment model with first-order elimination, parameterised with `CL` and `V` | structure | Methods, "Pharmacokinetic analysis" |
| `Vi = tvV * exp(etaV)`, `CLi = tvCL * exp(etaCL)` (exponential IIV) | structure | Methods, "Pharmacokinetic analysis" (displayed equation) |
| `AUCinf = Dose / CL` | structure | Methods, "Pharmacokinetic analysis" (displayed equation) |
| BSA entered on `CL` and `V`, normalised to the median 0.81 m^2 | structure | Methods, "Pharmacokinetic analysis"; Results, "Population pharmacokinetic analysis"; Supplemental Table 6 |
| `lcl` (`tvCL`) | 4.42 L/h | Supplemental Table 6 (RSE 1.33%; bootstrap 4.41) |
| `lvc` (`tvV`) | 14.4 L | Supplemental Table 6 (RSE 0.91%; bootstrap 14.3) |
| `e_bsa_cl` (`dCLdBSA`) | 1.14 | Supplemental Table 6 (RSE 2.20%; bootstrap 1.14) |
| `e_bsa_vc` (`dVdBSA`) | 1.33 | Supplemental Table 6 (RSE 1.46%; bootstrap 1.34) |
| `etalcl` variance | 0.0572 | Supplemental Table 6 caption ("the variance from the Omega matrix for hCL is 0.0572") |
| `etalvc` variance | 0.0279 | Supplemental Table 6 caption ("and for hV is 0.0279") |
| `propSd` | fixed(0) | **Not reported.** Methods states only that "A multiplicative random error model ... [was] used"; no sigma appears in the paper or in either supplement revision. See Errata. |
| Therapeutic range 14.8-24.6 mg*h/L | validation target | Methods, "Busulfan dosing and sampling schedule"; Supplemental Table 7 footnote |
| Nominal first-dose sampling times | validation input | Supplemental Table 2 |
| Typical values by age subgroup | falsifier | Table 3 |
| C.V. of individual clearances by normalisation | falsifier | Figure 2 caption |
| Simulated AUCinf by dosing strategy | falsifier | Supplemental Table 7 |
| BSA-banded infant dosing table | validation input | Table 1 |

## Reading Phoenix's `dCLdBSA` / `dVdBSA` as power exponents

Supplemental Table 6 names the two covariate coefficients `dCLdBSA` and `dVdBSA`
without printing the equation that consumes them. Phoenix NLME can build a
continuous-covariate effect as a power, an exponential, or a linear form, so the
form has to be settled from the paper's own numbers rather than assumed.

Table 3 provides the discriminator. It reports separate one-compartment fits to
three nested age subgroups, each with `CL` in both L/h and L/(h*m^2) and `V` in
both L and L/m^2. The ratio of the two columns is the subgroup's effective BSA,
which lets each subgroup be pushed through all three candidate forms and scored
against the reported values.

```{r covariate-form-falsifier}
tab3 <- tibble::tribble(
  ~subgroup,   ~n,   ~cl_lh, ~cl_lhm2, ~v_l,  ~v_lm2,
  "All",       328L, 4.43,   5.47,     14.5,  17.8,
  "<=3 years", 108L, 2.12,   4.95,     6.24,  14.5,
  "<=2 years",  88L, 1.92,   4.74,     5.70,  14.0,
  "<=1 year",   46L, 1.46,   4.27,     4.55,  13.3
) |>
  # Effective BSA of each subgroup, from the two ways the same fit is reported.
  mutate(BSA = (cl_lh / cl_lhm2 + v_l / v_lm2) / 2)

tvcl <- 4.42; tvv <- 14.4; bsa_ref <- 0.81
e_cl <- 1.14; e_v <- 1.33

forms <- tab3 |>
  mutate(
    cl_power = tvcl * (BSA / bsa_ref)^e_cl,
    cl_lin   = tvcl * (1 + e_cl * (BSA - bsa_ref)),
    cl_exp   = tvcl * exp(e_cl * (BSA - bsa_ref)),
    v_power  = tvv  * (BSA / bsa_ref)^e_v,
    v_lin    = tvv  * (1 + e_v  * (BSA - bsa_ref)),
    v_exp    = tvv  * exp(e_v  * (BSA - bsa_ref))
  )

pct <- function(model, reported) round(100 * (model - reported) / reported, 1)

forms |>
  transmute(
    subgroup,
    `Effective BSA (m^2)` = round(BSA, 3),
    `CL power (%)` = pct(cl_power, cl_lh),
    `CL linear (%)` = pct(cl_lin, cl_lh),
    `CL exponential (%)` = pct(cl_exp, cl_lh),
    `V power (%)` = pct(v_power, v_l),
    `V linear (%)` = pct(v_lin, v_l),
    `V exponential (%)` = pct(v_exp, v_l)
  ) |>
  knitr::kable(
    caption = paste(
      "Percent error of each candidate covariate form against the Table 3",
      "subgroup fits. The power form is the only one that reproduces the",
      "reported values; linear and exponential are off by tens of percent",
      "in the youngest subgroups."
    )
  )
```

The power form reproduces `V` to within about 1% in every subgroup and `CL` to
within 1% in the two larger subgroups; the linear and exponential forms miss by
14% to 39% in the infants. The residual `CL` over-prediction in the youngest
subgroup is the paper's own central finding -- infants have a lower BSA-scaled
clearance than the population power model predicts, which is precisely why the
authors propose a separate banded dosing table below a BSA of 0.5 m^2.

A second, independent check: at the reference BSA of 0.81 m^2 the packaged model
gives `CL` = 4.42 L/h, i.e. 4.42 / 0.81 = `r round(4.42 / 0.81, 2)` L/(h*m^2)
against the 5.47 L/(h*m^2) of Table 3, and `V` = 14.4 L, i.e.
14.4 / 0.81 = `r round(14.4 / 0.81, 1)` L/m^2 against the reported 17.8 L/m^2.

# Virtual cohort

The original patient-level data are not publicly available ("No datasets were
generated or analysed during the current study"). The cohort below is a virtual
population built to match the demographics of Balis 2025 Table 2.

Age is drawn from a piecewise-linear quantile function anchored on the three
facts the paper reports (minimum 0.11 years, 27% under 2 years, median
5.9 years, maximum 28.1 years). Weight and height are taken from CDC 50th
percentile growth references at the sampled age with lognormal scatter, and BSA
is computed with the Mosteller formula. The paper does not state which BSA
formula was used; Mosteller is the discriminator-free choice because the model
only ever sees BSA through the ratio BSA / 0.81.

```{r cohort}
set.seed(20250213)
n_sub <- 200L

# CDC 50th percentile weight and stature for age, averaged across sexes.
growth <- tibble::tribble(
  ~age,  ~wt,  ~ht,
  0.10,   4.4,  54,
  0.25,   6.0,  60,
  0.50,   7.6,  66,
  1.00,   9.6,  75,
  2.00,  12.2,  86,
  3.00,  14.3,  95,
  4.00,  16.3, 102,
  5.00,  18.3, 109,
  6.00,  20.6, 115,
  7.00,  23.0, 121,
  8.00,  25.6, 127,
 10.00,  32.0, 138,
 12.00,  40.5, 149,
 14.00,  50.0, 162,
 16.00,  58.0, 168,
 18.00,  63.0, 170,
 21.00,  68.0, 171,
 28.10,  75.0, 171
)

# Age quantile anchors from Balis 2025 Table 2 and Results.
age_p   <- c(0.000, 0.135, 0.270, 0.500, 0.750, 0.950, 1.000)
age_q   <- c(0.110, 0.750, 2.000, 5.900, 11.50, 18.50, 28.10)

cohort <- tibble::tibble(
  id  = seq_len(n_sub),
  AGE = stats::approx(age_p, age_q, xout = stats::runif(n_sub))$y
) |>
  mutate(
    wt_med = stats::approx(growth$age, growth$wt, xout = AGE, rule = 2)$y,
    ht_med = stats::approx(growth$age, growth$ht, xout = AGE, rule = 2)$y,
    WT     = wt_med * exp(stats::rnorm(n_sub, 0, 0.20)),
    HT     = ht_med * exp(stats::rnorm(n_sub, 0, 0.045)),
    BSA    = sqrt(HT * WT / 3600)
  ) |>
  select(id, AGE, WT, HT, BSA)

tibble::tibble(
  Characteristic = c("Age (years)", "Weight (kg)", "BSA (m^2)",
                     "Fraction < 2 years", "Fraction with BSA < 0.5 m^2"),
  Simulated = c(
    sprintf("%.1f (%.2f to %.1f)", median(cohort$AGE), min(cohort$AGE), max(cohort$AGE)),
    sprintf("%.1f (%.1f to %.0f)", median(cohort$WT),  min(cohort$WT),  max(cohort$WT)),
    sprintf("%.2f (%.2f to %.2f)", median(cohort$BSA), min(cohort$BSA), max(cohort$BSA)),
    sprintf("%.0f%%", 100 * mean(cohort$AGE < 2)),
    sprintf("%.0f%%", 100 * mean(cohort$BSA < 0.5))
  ),
  `Balis 2025` = c(
    "5.9 (0.11 to 28.1)", "20.5 (3.4 to 126)", "0.81 (0.21 to 2.42)",
    "27%", "21% (70 of 328)"
  )
) |>
  knitr::kable(
    caption = "Virtual cohort demographics vs Balis 2025 Table 2 (median (range))."
  )
```

# Simulation

Each subject receives a single first dose on each of the two schedules the paper
used: 3.2 mg/kg infused over 3 h (daily schedule) and 0.8 mg/kg infused over 2 h
(q6h schedule). The two arms use disjoint subject-ID ranges.

```{r events}
make_arm <- function(cohort, mg_per_kg, inf_dur, tmax, schedule, id_offset) {
  sub <- cohort |>
    mutate(id = id + id_offset, amt = mg_per_kg * WT, schedule = schedule)

  doses <- sub |>
    transmute(id, AGE, WT, HT, BSA, schedule, time = 0, amt,
              rate = amt / inf_dur, evid = 1L, cmt = "central")

  obs <- sub |>
    select(id, AGE, WT, HT, BSA, schedule) |>
    tidyr::crossing(time = seq(0, tmax, by = 0.25)) |>
    mutate(amt = 0, rate = 0, evid = 0L, cmt = "central")

  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}

events <- bind_rows(
  make_arm(cohort, 3.2, 3, 24, "Daily 3.2 mg/kg over 3 h",   0L),
  make_arm(cohort, 0.8, 2,  6, "q6h 0.8 mg/kg over 2 h",  1000L)
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))

mod <- readModelDb("Balis_2025_busulfan")
sim <- rxode2::rxSolve(
  mod, events = events,
  keep = c("schedule", "AGE", "WT", "BSA")
) |>
  as.data.frame()

# rxSolve silently drops subjects whose events it cannot expand; assert instead.
stopifnot(dplyr::n_distinct(sim$id) == 2L * n_sub)
```

## Concentration-time profiles

```{r figure-profiles}
sim |>
  group_by(schedule, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  # Drop the pre-dose zero so the log scale has something to draw.
  filter(Q05 > 0) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~schedule, scales = "free_x") +
  scale_y_log10() +
  labs(
    x = "Time after start of infusion (h)",
    y = "Busulfan concentration (mg/L)",
    title = "Simulated first-dose busulfan profiles",
    caption = "Median and 5th-95th percentile, 200 virtual subjects per schedule."
  )
```

## Replicating Figure 2: clearance scaled to body weight vs BSA

Figure 2 of Balis 2025 plots individual clearances against age, scaled to body
weight (panel A) and to BSA (panel B), and its caption reports the coefficient of
variation of `CLi` under three normalisations. Those three C.V.s are a direct
test of the covariate model: a BSA power exponent close to 1 should collapse the
spread of `CLi`/BSA to little more than the IIV alone, while leaving `CLi`/WT and
unnormalised `CLi` much wider.

```{r figure-2}
cl_i <- sim |>
  distinct(id, schedule, AGE, WT, BSA, cl) |>
  filter(schedule == "Daily 3.2 mg/kg over 3 h") |>
  mutate(cl_wt = cl / WT, cl_bsa = cl / BSA)

cv <- function(x) 100 * stats::sd(x) / mean(x)

tibble::tibble(
  Normalisation = c("CLi / BSA  [L/(h*m^2)]", "CLi / weight  [L/(h*kg)]",
                    "CLi, not normalised  [L/h]"),
  `Simulated C.V.` = sprintf("%.0f%%", c(cv(cl_i$cl_bsa), cv(cl_i$cl_wt), cv(cl_i$cl))),
  `Balis 2025 Figure 2` = c("23%", "28%", "62%")
) |>
  knitr::kable(
    caption = "Coefficient of variation of individual clearance under three normalisations."
  )

cl_i |>
  select(AGE, `CLi / weight [L/(h*kg)]` = cl_wt, `CLi / BSA [L/(h*m^2)]` = cl_bsa) |>
  tidyr::pivot_longer(-AGE, names_to = "panel", values_to = "value") |>
  ggplot(aes(AGE, value)) +
  geom_point(shape = 1, alpha = 0.7) +
  facet_wrap(~panel, scales = "free_y") +
  labs(
    x = "Age (years)", y = "Individual clearance",
    title = "Replicates Figure 2 of Balis 2025",
    caption = paste(
      "CLi scaled to BSA is flatter across the age span than CLi scaled to",
      "body weight."
    )
  )
```

# PKNCA validation

NCA is run twice. The primary run uses the full 0.25 h simulation grid, which
resolves the infusion phase and therefore measures the model rather than the
sampling design. A second run uses only the nominal first-dose sampling times of
Supplemental Table 2 -- 3, 3.25, 3.5, 4, 5, 6 and 8 h on the daily schedule and
2, 2.25, 2.5, 3, 4, 5 and 6 h on the q6h schedule -- to quantify what the paper's
own sampling design costs a trapezoidal AUC.

```{r pknca}
dose_df <- events |>
  filter(evid == 1L) |>
  select(id, time, amt, schedule)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

run_nca <- function(conc_df) {
  # Guarantee a time-zero record; busulfan is absent before the first infusion.
  conc_df <- bind_rows(
    conc_df,
    conc_df |> distinct(id, schedule) |> mutate(time = 0, Cc = 0)
  ) |>
    distinct(id, schedule, time, .keep_all = TRUE) |>
    arrange(id, schedule, time)

  conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | schedule + id,
                               concu = "mg/L", timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | schedule + id, doseu = "mg")
  as.data.frame(
    PKNCA::pk.nca(
      PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
    )$result
  )
}

sim_conc <- sim |>
  filter(!is.na(Cc)) |>
  transmute(id, schedule = as.character(schedule), time = round(time, 4), Cc)

nca_tbl <- run_nca(sim_conc)

# The paper's nominal first-dose sampling times (Supplemental Table 2).
sample_times <- bind_rows(
  tibble::tibble(schedule = "Daily 3.2 mg/kg over 3 h",
                 time = c(3, 3.25, 3.5, 4, 5, 6, 8)),
  tibble::tibble(schedule = "q6h 0.8 mg/kg over 2 h",
                 time = c(2, 2.25, 2.5, 3, 4, 5, 6))
)

sim_conc_nominal <- sim_conc |>
  inner_join(mutate(sample_times, time = round(time, 4)),
             by = c("schedule", "time"))
stopifnot(nrow(sim_conc_nominal) == 7L * 2L * n_sub)

nca_tbl_nominal <- run_nca(sim_conc_nominal)
```

## Per-subject identity check: AUCinf = Dose / CL

The paper computes every simulated exposure as `AUCinf = Dose / CL`. For a
one-compartment linear model with intravenous input that identity is exact, so it
is the sharpest available check that PKNCA and the packaged model agree. It is
asserted per subject rather than on a cohort median, which would hide
compensating errors.

```{r auc-identity}
check_identity <- function(tbl) {
  tbl |>
    filter(PPTESTCD == "aucinf.obs") |>
    select(schedule, id, aucinf = PPORRES) |>
    mutate(id = as.integer(as.character(id))) |>
    inner_join(
      sim |> distinct(id, schedule, cl) |>
        mutate(schedule = as.character(schedule)) |>
        inner_join(dose_df |> select(id, amt), by = "id"),
      by = c("id", "schedule")
    ) |>
    mutate(
      auc_theory = amt / cl,
      pct_err    = 100 * (aucinf - auc_theory) / auc_theory
    )
}

auc_check <- check_identity(nca_tbl)
summary(auc_check$pct_err)
stopifnot(max(abs(auc_check$pct_err)) < 1)
```

On the full simulation grid the largest per-subject deviation is
`r sprintf("%.2f%%", max(abs(auc_check$pct_err)))`: PKNCA's log-linear
down-slope integration is exact for a mono-exponential decline, so the residual
is only the linear-trapezoid discretisation of the rising infusion phase.

## What the published sampling design costs a trapezoidal AUC

Balis 2025 notes that AUCs "recalculated using the trapezoidal rule in Phoenix's
NCA module ... were in close agreement (mean difference 1.7%) with the clinically
reported values", the latter being the one-compartment model fits. The nominal
sampling schedule takes its first sample at the end of infusion, so the entire
rising phase is spanned by a single trapezoid across a concave curve and the
trapezoidal AUC must under-read. Repeating the identity check on the nominal
times alone isolates that effect.

```{r nominal-sampling-cost}
check_identity(nca_tbl_nominal) |>
  group_by(schedule) |>
  summarise(
    `Mean difference (%)`   = round(mean(pct_err), 1),
    `Median difference (%)` = round(median(pct_err), 1),
    .groups = "drop"
  ) |>
  rename(Schedule = schedule) |>
  knitr::kable(
    caption = paste(
      "Trapezoidal AUCinf from the nominal sampling times vs the model's",
      "Dose / CL, by schedule. Balis 2025 reports a mean difference of 1.7%",
      "between its trapezoidal and model-fitted AUCs."
    )
  )
```

The bias is negative and larger on the daily schedule, whose 3 h infusion leaves
a longer unsampled rise than the q6h schedule's 2 h infusion -- the expected
direction and ordering. The magnitude exceeds the paper's 1.7% because the
patients' *actual* sample times were used in the published analysis rather than
the nominal times tabulated in Supplemental Table 2, and because Phoenix's
handling of the infusion interval need not match PKNCA's. This block is a
directional check on the sampling design, not a reproduction of the 1.7% figure.

## Comparison against published NCA

Balis 2025 reports the mean and standard deviation of AUCinf after the first dose
separately for the two schedules: 16.2 +/- 5.1 mg*h/L on the daily schedule
(n = 154) and 4.12 +/- 1.23 mg*h/L on the q6h schedule (n = 174).

```{r nca-compare}
sim_summary <- nca_tbl |>
  filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  group_by(schedule, PPTESTCD) |>
  summarise(PPORRES = mean(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

published <- tibble::tribble(
  ~schedule,                    ~aucinf.obs,
  "Daily 3.2 mg/kg over 3 h",   16.2,
  "q6h 0.8 mg/kg over 2 h",      4.12
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_summary,
  reference = published,
  by        = "schedule",
  params    = "aucinf.obs",
  units     = c(aucinf.obs = "mg*h/L"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Mean simulated AUCinf after the first dose vs the observed NCA means of",
    "Balis 2025 Results, 'Noncompartmental Analysis'.",
    "* differs from reference by more than 20%."
  )
)
```

The published daily-schedule cohort had a median age of 9.02 years and the q6h
cohort 4.11 years, whereas the virtual cohort here draws both arms from the
pooled population median of 5.9 years. Because the dose is scaled to body weight
while clearance scales with BSA, the weight-to-BSA ratio rises with age and so
does AUCinf per mg/kg; the pooled-age cohort therefore biases the daily arm low
and the q6h arm high relative to the published, age-separated means. Evaluating
the packaged model at each schedule's own published median age removes the
confound:

```{r schedule-median-age}
typical_at_age <- function(age, mg_per_kg) {
  wt  <- stats::approx(growth$age, growth$wt, xout = age, rule = 2)$y
  ht  <- stats::approx(growth$age, growth$ht, xout = age, rule = 2)$y
  bsa <- sqrt(ht * wt / 3600)
  tibble::tibble(
    `Median age (years)` = age,
    `Weight (kg)`        = round(wt, 1),
    `BSA (m^2)`          = round(bsa, 2),
    `Dose (mg)`          = round(mg_per_kg * wt, 1),
    `Model AUCinf (mg*h/L)` = round(mg_per_kg * wt / (tvcl * (bsa / bsa_ref)^e_cl), 1)
  )
}

bind_rows(
  typical_at_age(9.02, 3.2) |> mutate(Schedule = "Daily, 3.2 mg/kg", .before = 1),
  typical_at_age(4.11, 0.8) |> mutate(Schedule = "q6h, 0.8 mg/kg",   .before = 1)
) |>
  mutate(`Balis 2025 observed mean (mg*h/L)` = c(16.2, 4.12)) |>
  knitr::kable(
    caption = paste(
      "Typical-subject AUCinf at each schedule's published median age",
      "(Balis 2025 Results, 'Population pharmacokinetic analysis')."
    )
  )
```

At its own published median age the daily schedule lands within 2% of the
observed mean. The q6h typical subject remains about 13% low, for two reasons the
paper documents but this typical-subject calculation cannot absorb: 44 of the 174
q6h patients received 1 mg/kg rather than 0.8 mg/kg (a dose-weighted mean of
0.85 mg/kg, which alone accounts for roughly half the gap), and the q6h cohort is
the younger of the two, where the BSA power model over-predicts clearance.

# Replicating Supplemental Table 7: simulated AUCinf by dosing strategy

This is the paper's headline result. Three dosing strategies are compared:
body-weight-scaled dosing as actually practised, a flat 100 mg/m^2, and
100 mg/m^2 for patients with BSA >= 0.5 m^2 combined with the BSA-banded infant
dosing table of Table 1 for smaller patients. Exposures are computed exactly as
the paper computes them, `AUCinf = Dose / CLi`.

```{r suppl-table-7}
combined_lbl <- "100 mg/m^2 (BSA >= 0.5) + dosing table (BSA < 0.5)"

# Table 1 of Balis 2025, daily schedule column.
banded_dose <- function(bsa) {
  dplyr::case_when(
    bsa < 0.25 ~ 18,
    bsa < 0.30 ~ 20,
    bsa < 0.35 ~ 25,
    bsa < 0.40 ~ 32,
    bsa < 0.45 ~ 40,
    bsa < 0.50 ~ 50,
    TRUE       ~ 100 * bsa
  )
}

dosing <- cl_i |>
  select(id, AGE, WT, BSA, cl) |>
  mutate(
    # Methods: 3.2 mg/kg, or 4 mg/kg for children > 10 kg and < 48 months.
    dose_kg    = ifelse(WT > 10 & AGE < 4, 4.0, 3.2) * WT,
    dose_bsa   = 100 * BSA,
    dose_table = banded_dose(BSA)
  ) |>
  tidyr::pivot_longer(
    c(dose_kg, dose_bsa, dose_table),
    names_to = "strategy", values_to = "dose"
  ) |>
  mutate(
    strategy = factor(
      strategy,
      levels = c("dose_kg", "dose_bsa", "dose_table"),
      labels = c("Dose per kg", "100 mg/m^2", combined_lbl)
    ),
    auc = dose / cl
  )

lo <- 14.8; hi <- 24.6

summarise_strategy <- function(dat) {
  dat |>
    group_by(strategy) |>
    summarise(
      `In range (%)` = round(100 * mean(auc >= lo & auc <= hi), 1),
      `Below (%)`    = round(100 * mean(auc < lo), 1),
      `Above (%)`    = round(100 * mean(auc > hi), 1),
      `Median AUC (mg*h/L)` = round(median(auc), 1),
      .groups = "drop"
    )
}

published_st7 <- tibble::tribble(
  ~Subset,          ~strategy,      ~in_range, ~median_auc,
  "All patients",   "Dose per kg",       50.6,        15.8,
  "All patients",   "100 mg/m^2",        71.3,        18.0,
  "All patients",   combined_lbl,        75.3,        17.6,
  "BSA < 0.5 m^2",  "Dose per kg",       51.4,        15.3,
  "BSA < 0.5 m^2",  "100 mg/m^2",        57.1,        21.7,
  "BSA < 0.5 m^2",  combined_lbl,        75.7,        19.5
)

bind_rows(
  summarise_strategy(dosing) |> mutate(Subset = "All patients", .before = 1),
  summarise_strategy(filter(dosing, BSA < 0.5)) |>
    mutate(Subset = "BSA < 0.5 m^2", .before = 1)
) |>
  mutate(strategy = as.character(strategy)) |>
  # Join by key, never by row position -- summarise() and tribble() need not agree on order.
  inner_join(published_st7, by = c("Subset", "strategy")) |>
  rename(
    Strategy                   = strategy,
    `Published in range (%)`   = in_range,
    `Published median AUC`     = median_auc
  ) |>
  knitr::kable(
    caption = paste(
      "Simulated AUCinf by dosing strategy vs Balis 2025 Supplemental Table 7.",
      "Therapeutic range 14.8-24.6 mg*h/L."
    )
  )
```

```{r figure-3}
dosing |>
  filter(strategy != "100 mg/m^2") |>
  ggplot(aes(BSA, auc)) +
  geom_hline(yintercept = c(lo, hi), colour = "darkgreen") +
  geom_point(shape = 1, alpha = 0.7) +
  facet_wrap(~strategy) +
  scale_y_log10() +
  labs(
    x = expression(paste("Body surface area (", m^2, ")")),
    y = "Simulated AUCinf (mg*h/L)",
    title = "Replicates Figure 3 of Balis 2025",
    caption = paste(
      "Green lines are the bounds of the therapeutic range",
      "(14.8 and 24.6 mg*h/L)."
    )
  )
```

Across all patients the reproduction is close. Weight-scaled dosing leaves
roughly 40% of the population subtherapeutic after the first dose (published
41.5%), and moving to a flat 100 mg/m^2 lifts the in-range fraction to 70.5%
against the published 71.3% with a median AUCinf of 18.1 against 18.0 mg*h/L.

The `BSA < 0.5 m^2` rows are the ones that do not reproduce, and they fail in a
single consistent direction: every simulated exposure in that subset is low. This
is not a transcription problem, it is the paper's own central finding showing
through. Table 3 reports a clearance of 4.27 L/(h*m^2) in patients <= 1 year,
whereas the BSA power model evaluated at that subgroup's BSA predicts about
4.84 L/(h*m^2) -- a 13% over-prediction of infant clearance, which maps to a
roughly 13% under-prediction of AUCinf at a fixed dose. The observed gap in the
`Dose per kg` row is 12.9 against 15.3 mg*h/L, or -16%, and the two banded-table
rows are displaced similarly. The authors sized the banded dosing table from
their infants' *post hoc* individual clearances, which carry those systematically
negative `etalcl` values; a virtual cohort drawing `etalcl` from a mean-zero
distribution cannot reproduce that, and so it under-doses the smallest patients
relative to the published simulation. The consequence is that the combined
strategy scores slightly *below* the flat 100 mg/m^2 strategy here (68.0% vs
70.5% in range), inverting the published ordering (75.3% vs 71.3%). The
qualitative conclusion of the paper -- that BSA-scaled dosing plus an infant
table beats weight-scaled dosing -- reproduces; its precise margin over flat
BSA-scaled dosing does not, because that margin is carried by information in the
individual clearances that the published covariate model does not contain.

# Assumptions and deviations

* **Residual error is not reported and is encoded as `propSd <- fixed(0)`.**
  Balis 2025 Methods states only that "A multiplicative random error model and
  the FOCE-ELS ... algorithm ... were used for all population pharmacokinetic
  model fits"; no sigma estimate appears in the paper, in Supplemental Table 6,
  or in either revision of the electronic supplement. The magnitude was not
  invented. Supplemental Figure 3B (measured vs individual-predicted
  concentrations on log-log axes) shows a visibly tight band, so the residual
  error is small, but the figure does not support a defensible point estimate.
  Every validation in this vignette is residual-error-free by construction --
  the paper's own exposures are computed as `Dose / CL` -- so the omission does
  not affect any comparison above. Users who need to simulate observed
  concentrations should supply their own `propSd`.
* **The covariate form is inferred, not printed.** Supplemental Table 6 names
  `dCLdBSA` and `dVdBSA` but never writes the equation. The power form
  `theta * (BSA / 0.81)^d` was selected by scoring all three Phoenix continuous
  covariate forms against the Table 3 subgroup fits (see "Reading Phoenix's
  `dCLdBSA` / `dVdBSA` as power exponents"), where it is the only form that
  reproduces the reported values.
* **BSA formula.** The paper abstracts weight, height and BSA from the medical
  record without naming a BSA equation. The virtual cohort uses Mosteller,
  `sqrt(height_cm * weight_kg / 3600)`. The model consumes BSA only through
  `BSA / 0.81`, so a different BSA equation shifts the cohort, not the model.
* **Growth references.** Weight and height for age come from CDC 50th percentile
  growth references averaged across sexes, with lognormal scatter (SD 0.20 on
  log weight, 0.045 on log height) chosen so the resulting weight and BSA ranges
  bracket those of Table 2. Sex was not a covariate in the model (Balis 2025
  Results: median CLi 5.65 vs 5.38 L/(h*m^2) in males and females, not
  statistically significantly different), so the cohort is not sex-stratified.
* **Age distribution.** Drawn from a piecewise-linear quantile function anchored
  on the published minimum, median, maximum and the "27% under 2 years" figure.
  The intermediate quantiles are an assumption.
* **Single first dose.** The packaged model is the day-1 fit. Balis 2025 reports
  that clearance falls by a median of 11% from day 1 to day 2 (Supplemental
  Table 5) and attributes it to glutathione depletion, but does not fit a
  time-dependent clearance model, so no such term is encoded here. Simulations
  of a full 4-day conditioning course will therefore under-predict later-day
  exposure. A busulfan model that does carry an explicit time-decaying
  elimination rate constant is available as
  `modellib("Schreib_2024_busulfan")`.
* **Cohort size.** 200 virtual subjects per arm rather than the 328 real
  patients, per the nlmixr2lib per-arm cap. Published percentages in
  Supplemental Table 7 are over 328 patients and 70 patients with BSA
  < 0.5 m^2; the simulated percentages carry correspondingly wider Monte Carlo
  noise, particularly in the BSA < 0.5 m^2 subset.
* **"Dose per kg" row of Supplemental Table 7.** The paper's footnote describes
  that row as "the actual AUCinf achieved after a dose scaled to body weight",
  i.e. observed NCA rather than a model simulation. It is reproduced here by
  simulation for a like-for-like comparison across strategies, which is why it
  is the row most sensitive to the virtual cohort's weight distribution.
