---
title: "Rifampicin (Barnett 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rifampicin (Barnett 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Barnett_2018_rifampicin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Barnett_2018_rifampicin"))$description`
* Article: <https://doi.org/10.1002/cpt.983>

# Population

Barnett 2018 refit the rifampicin popPK model to plasma rifampicin data from the Lai et al. 2016 clinical drug-drug-interaction (DDI) study in 12 healthy adult male volunteers (SLCO1B1 c.521 T>C wildtype only; no OATP1B1*5 / *15 carriers). Each subject received a single 600 mg oral rifampicin dose at the start of occasion 1 (rifampicin alone) and at the start of occasion 3 (rifampicin co-administered with 5 mg rosuvastatin), with a 7-day washout between occasions and rifampicin omitted on occasion 2. Per-subject demographic distributions (age range, body weight range, race/ethnicity, region) were not tabulated in Barnett 2018; the Methods section cites the source clinical study (Lai Y et al. Pharmacol Res Perspect 2016) for those baseline characteristics. The 276 rifampicin plasma samples used for modeling purposes were collected over the 24 h post-dose window on occasion 1 (n = 144) and occasion 3 (n = 132).

The same demographic context is available programmatically via `readModelDb("Barnett_2018_rifampicin")$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()` entry in `inst/modeldb/specificDrugs/Barnett_2018_rifampicin.R`. The table below collects the most important entries in one place for review.

| Equation / parameter | Value           | Source location                                       |
|----------------------|-----------------|-------------------------------------------------------|
| `lka`                | log(2.09)       | Barnett 2018 Table 1, RIF row 'ka (1/h)'              |
| `lcl`                | log(3.97)       | Barnett 2018 Table 1, RIF row 'CL (L/h)'              |
| `lvc`                | log(24.7)       | Barnett 2018 Table 1, RIF row 'V (L)'                 |
| `lmtt`               | log(0.74)       | Barnett 2018 Table 1, RIF row 'MTT (h)'               |
| `lnn`                | log(8.63)       | Barnett 2018 Table 1, RIF row 'n'                     |
| IIV ka / CL / V / MTT / n | 46% / 30.7% / 19.5% / 69.1% / 24.3% CV | Barnett 2018 Table 1, IIV column                      |
| IOV ka / V / MTT     | 49.2% / 6.6% / 47.6% CV | Barnett 2018 Table 1, IOV column                      |
| `propSd`             | 0.313 (= 31.3%) | Barnett 2018 Table 1, RIF row 'r prop (%)'            |
| `addSd`              | 0.01 umol/L FIXED | Barnett 2018 Table 1, RIF row 'r add (uM)'          |
| Structural form (1-cmt + transit absorption) | n/a | Barnett 2018 Results, Rifampicin popPK model section; structural model adapted from Wilkins JJ et al. AAC 2008;52(6):2138-2148. |

The transit-chain input rate uses the Savic 2007 / Wilkins 2008 analytical form via rxode2's `transit(n, mtt)` function; see the model file's `model()` block for the exact implementation. Barnett 2018 Table 1 footnote 'a' notes that the NONMEM covariance step for this fit failed, so per-parameter standard errors are not reported by the source.

# Virtual cohort

The packaged model is parameterised for a typical adult male healthy volunteer. Stochastic VPC-style simulations use the published IIV and IOV variances to reproduce between-subject and between-occasion spread; deterministic typical-value simulations zero out the random effects via `rxode2::zeroRe()`.

```{r cohort}
set.seed(254)

mod <- readModelDb("Barnett_2018_rifampicin")

n_sub <- 50L

make_cohort_events <- function(n, time_grid = seq(0.1, 24, by = 0.1), occ = 1) {
  doses <- data.frame(
    id   = seq_len(n),
    time = 0,
    evid = 1L,
    amt  = 600,
    cmt  = "depot",
    OCC  = occ
  )
  obs <- data.frame(
    id   = rep(seq_len(n), each = length(time_grid)),
    time = rep(time_grid, times = n),
    evid = 0L,
    amt  = NA_real_,
    cmt  = "Cc",
    OCC  = occ
  )
  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time)
}

events <- make_cohort_events(n_sub, occ = 1)
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(mod, events = events, keep = c("OCC"))
sim_df <- as.data.frame(sim)

mod_typical <- rxode2::zeroRe(mod)
sim_typical <- rxode2::rxSolve(mod_typical, events = make_cohort_events(1L, occ = 1)) |>
  as.data.frame()
```

# Typical-value rifampicin plasma profile

```{r plot-typical}
ggplot(sim_typical, aes(time, Cc)) +
  geom_line(linewidth = 1) +
  labs(x = "Time (h)", y = "Rifampicin Cc (umol/L)",
       title = "Single 600 mg oral rifampicin in a typical adult male",
       caption = "Typical-value (no IIV / no residual error) prediction; Barnett 2018 popPK parameters.") +
  theme_minimal()
```

# Visual predictive check (50 simulated subjects)

```{r plot-vpc}
vpc <- sim_df |>
  dplyr::filter(!is.na(Cc), time > 0) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    q05 = quantile(Cc, 0.05, na.rm = TRUE),
    q50 = quantile(Cc, 0.50, na.rm = TRUE),
    q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(vpc, aes(time, q50)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.25) +
  geom_line(linewidth = 1) +
  labs(x = "Time (h)", y = "Rifampicin Cc (umol/L)",
       title = "Simulated VPC-style envelope for 50 healthy adults",
       caption = "Median + 90% PI from a stochastic simulation. Compare visually to Barnett 2018 Supplementary Figure S7.") +
  theme_minimal()
```

# PKNCA validation

```{r pknca}
sim_nca <- sim_df |>
  dplyr::filter(!is.na(Cc), time > 0) |>
  dplyr::select(id, time, Cc, OCC)
sim_nca$treatment <- "RIF 600 mg PO"

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, OCC)
dose_df$treatment <- "RIF 600 mg PO"

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "umol/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "mg")

intervals <- data.frame(
  start       = 0,
  end         = Inf,
  cmax        = TRUE,
  tmax        = TRUE,
  aucinf.obs  = TRUE,
  half.life   = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- PKNCA::pk.nca(nca_data)

nca_summary <- summary(nca_res)
knitr::kable(nca_summary, caption = "Simulated NCA parameters for the 50-subject virtual cohort (single 600 mg PO RIF).")
```

# Comparison against the source paper

Barnett 2018 Table 1 does not tabulate observed NCA values for rifampicin; only the structural popPK estimates are reported, and the visual-predictive-check fit is shown in Supplementary Figure S7. A rough sanity comparison against externally published rifampicin PK in healthy adults (e.g., Acocella 1978, Burman 2001) places the expected Cmax after a single 600 mg oral dose in the 10-20 mg/L (12-24 umol/L) range and the half-life in the 2-5 h range; the simulated typical-value Cmax of ~24 umol/L and half-life of ~5 h fall within those expected envelopes.

# Assumptions and deviations

* **No SE on per-parameter estimates.** Barnett 2018 Table 1 footnote 'a' notes that the NONMEM covariance step for the rifampicin model failed; the point estimates were reported but per-parameter standard errors were not. The model file encodes IIV / IOV variances as if estimated and does not propagate parameter-estimation uncertainty.
* **IOV occasion structure.** Barnett 2018 reports a single shared IOV variance per parameter (ka, V, MTT) across the rifampicin-dosing occasions. The model file encodes three OCC slots (OCC1, OCC2, OCC3) with the variance shared via NONMEM's `OMEGA BLOCK(1) SAME` idiom; OCC2 (rosuvastatin-only period) is included for completeness even though no rifampicin data was collected on that occasion in the source study.
* **Demographics.** Barnett 2018 does not tabulate per-subject body-weight / age / region data for the n = 12 cohort. The model file's `population$weight_range`, `population$age_range`, etc. mark these as not extracted.
* **Structural model carried from Wilkins 2008.** The 1-compartment + transit absorption form was selected by Barnett 2018 based on prior literature (Wilkins 2008); see `modellib('Wilkins_2008_rifampicin')` for the original DDMORE-curated implementation of that structural form. Barnett 2018 reports refit parameter point estimates, not the prior literature values.
