---
title: "Sutimlimab (Bartko 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sutimlimab (Bartko 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tibble)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Bartko_2018_sutimlimab"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Bartko_2018_sutimlimab"))$description`
* Article: <https://doi.org/10.1002/cpt.1111>

Sutimlimab (BIVV009 / TNT009; now marketed as Enjaymo) is a humanized monoclonal
IgG4 antibody directed against complement factor C1s. By binding C1s it blocks
the enzymatic propagation of the classical complement pathway (CP) while leaving
the alternative and lectin pathways, and the opsonic function of C1q, intact.
Bartko 2018 is the first-in-human trial: a phase I, double-blind, randomized,
placebo-controlled, single- (part A) and multiple- (part B) ascending-dose study
in healthy volunteers (NCT02502903).

The packaged model is the trial's **pharmacodynamic** model: a direct-effect
sigmoidal inhibitory Emax (Imax) relationship between serum sutimlimab
concentration and CP activity. Bartko 2018 first checked the
concentration-effect relationship for hysteresis, found none, and therefore
time-matched each individual serum concentration with the CP activity measured
in the same sample (Methods, "PK/PD").

There is **no packaged PK model**, because the paper does not contain one: the
sutimlimab pharmacokinetics are characterised by model-independent NCA only
(Table 2). See *Assumptions and deviations* below.

# Population

Bartko 2018 enrolled 64 healthy volunteers at a single centre (Department of
Clinical Pharmacology, Medical University of Vienna, Austria), of whom 48
received sutimlimab and 16 received placebo. Part A gave a single ~60-minute
i.v. infusion of 0.3, 1, 3, 10, 30, 60, or 100 mg/kg (n = 3 active for the two
lowest doses, n = 6 active for the rest, 3:1 active:placebo). Part B gave four
once-weekly ~60-minute i.v. infusions of 30 or 60 mg/kg (n = 6 active per dose
level) with a two-week follow-up.

Across the sutimlimab arms the median age was 32 years (range 19-59) in part A
and 27 years (22-41) in part B; 22 of 48 (45.8%) were female; 46 of 48 (95.8%)
were Caucasian and 2 (4.2%) African. Cohort mean body weights ran from 55 kg
(the 100 mg/kg cohort, which had a 58 kg protocol weight cap) to 79 kg; all
other cohorts were capped at 98 kg. Baseline CP activity was normal in every
subject (part A sutimlimab 97 +/- 14%, part B sutimlimab 94 +/- 18%). Baseline
demographics are Bartko 2018 Table 1.

The same demographics are exposed programmatically via the model's `population`
metadata:

```{r population}
pop <- rxode2::rxode(readModelDb("Bartko_2018_sutimlimab"))$population
str(pop, max.level = 1)
```

# Source trace

The per-parameter origin is recorded as in-file comments next to each `ini()`
entry in `inst/modeldb/specificDrugs/Bartko_2018_sutimlimab.R`. The table below
collects them in one place for review. All four typical values come from
Supplementary Table S3, "PK/PD parameters of BIVV009 and CP activity - parts A
and B"; the parenthesised figures in that table are relative standard errors of
the estimate (RSE%), i.e. parameter precision, and are **not** interindividual
variability.

| Equation / parameter | Value (RSE%) | Source location |
|----------------------|--------------|-----------------|
| `E0` (`le0 = log(94.8)`) | 94.8% (1.1) | Bartko 2018 Supplementary Table S3 |
| `Imax` (`limax = log(90.2)`) | 90.2 percentage points (1.1) | Bartko 2018 Supplementary Table S3; also Results, "PK/PD correlations" |
| `IC50` (`lic50 = log(6.2)`) | 6.2 ug/mL (27.5) | Bartko 2018 Supplementary Table S3; also Results, "PK/PD correlations" |
| `hill` (`lhill = log(2.4)`) | 2.4 (19.9) | Bartko 2018 Supplementary Table S3; also Abstract and Results |
| `addSd = fixed(0)` | not reported | Not tabulated anywhere in Bartko 2018 or its five supplementary files -- see *Assumptions and deviations* |
| Sigmoidal Imax equation `CPactivity = e0 - imax * C^hill / (ic50^hill + C^hill)` | n/a | Bartko 2018 Methods, "PK/PD" ("inhibitory Emax model"), Results "PK/PD correlations", and the fitted curve in Figure 5 |
| Direct effect (no effect compartment, no hysteresis) | n/a | Bartko 2018 Results, "PK/PD correlations" ("no delay was observed in CP activity"; concentrations and CP activity were time-matched) |
| Driving covariate `CP_SUTIMLIMAB_UGML` | ug/mL | Bartko 2018 Methods, "Pharmacokinetics" (validated immunoassay, Vela Laboratories) |
| IC90 = 15.5 ug/mL (cross-check, not a model parameter) | 15.5 ug/mL | Bartko 2018 Abstract and Results, "PK/PD correlations" |

## Internal consistency of the published parameter set

The paper reports an IC90 of 15.5 ug/mL separately from IC50 and the Hill
coefficient. Under the sigmoidal form, the concentration producing 90% of the
maximum effect is `IC50 * 9^(1/hill)`. That identity over-determines the
parameter set and validates all three transcriptions at once:

```{r ic90-identity}
ic50 <- 6.2
hill <- 2.4
ic90_implied <- ic50 * 9^(1 / hill)
c(published_IC90 = 15.5, implied_IC90 = round(ic90_implied, 3))
stopifnot(abs(ic90_implied - 15.5) < 0.05)
```

# Which Imax parameterisation?

Supplementary Table S3 reports `E0 (%) = 94.8` and `Imax (%) = 90.2` without
writing out the equation, which leaves two readings of how `Imax` enters:

* **additive** -- `Imax` is in the same units as `E0` (percentage points), so the
  saturating asymptote is `E0 - Imax = 4.6%`;
* **multiplicative** -- `Imax` is a fraction of `E0`, so the asymptote is
  `E0 * (1 - 0.902) = 9.29%`.

The paper's own fitted curve in Figure 5 settles it. Digitising the blue
"Predicted" line (300 dpi render of the published page; y-axis calibrated on the
0/50/100/150 major ticks, x-axis on the 1/10/100/1000/10000 decade ticks) gives
a curve that runs from 94.5% at its left end down to a high-concentration
plateau of 4.2-4.4%. That is a span of **90.3 percentage points**, which
reproduces `Imax = 90.2` directly and is only possible under the additive form.

```{r imax-parameterisation}
# Anchor points read off the published Figure 5 "Predicted" curve.
fig5 <- tibble::tribble(
  ~conc,   ~digitised,
     3.0,       85.88,
     5.0,       61.61,
     6.2,       50.84,
     9.9,       27.20,
    15.6,       13.60,
    29.9,        6.59,
    59.9,        4.92,
   100.0,        4.71,
   290.0,        4.39,
  1045.9,        4.39,
  2502.2,        4.39
)

e0 <- 94.8; imax <- 90.2
frac <- function(cc) cc^hill / (ic50^hill + cc^hill)

fig5 <- fig5 |>
  dplyr::mutate(
    additive       = e0 - imax * frac(conc),
    multiplicative = e0 * (1 - (imax / 100) * frac(conc))
  )

knitr::kable(
  fig5 |>
    dplyr::mutate(dplyr::across(c(digitised, additive, multiplicative), \(x) round(x, 2))) |>
    dplyr::rename(
      "Sutimlimab (ug/mL)"       = conc,
      "Figure 5 (digitised, %)"  = digitised,
      "Additive form (%)"        = additive,
      "Multiplicative form (%)"  = multiplicative
    ),
  caption = "Published Figure 5 fitted curve versus the two candidate Imax parameterisations."
)

# Root-mean-square disagreement with the published curve, over the anchors at
# or above 5 ug/mL (the left-most terminus of the drawn line is a rendering
# artefact and is excluded; see the narrative below).
keep <- fig5$conc >= 5
rmse <- c(
  additive       = sqrt(mean((fig5$digitised[keep] - fig5$additive[keep])^2)),
  multiplicative = sqrt(mean((fig5$digitised[keep] - fig5$multiplicative[keep])^2))
)
round(rmse, 3)

# The additive form must beat the multiplicative form by a wide margin. Both
# sides here are deterministic (fixed published constants against fixed
# digitised anchors), so an exact bound is appropriate.
stopifnot(
  rmse[["additive"]] < 1,
  rmse[["multiplicative"]] > 3,
  rmse[["multiplicative"]] > 4 * rmse[["additive"]]
)
```

Over the full digitised trace (403 sampled columns; the 349 at or above
5 ug/mL) the additive form reproduces the published curve with an RMSE of
0.35 percentage points and a maximum absolute error of 1.17, while the
multiplicative form is systematically 3-5 percentage points high with an RMSE of
4.32. The two forms are nearly indistinguishable below ~3 ug/mL, so the
left-most few pixels of the drawn line -- where the published curve terminates
at the lowest observed concentration and reads about 8 points above *both*
candidate forms -- cannot discriminate between them and is excluded above.

# Virtual cohort

The model is algebraic and carries no interindividual variability and no
residual error (see *Assumptions and deviations*), so it is fully
deterministic: one "subject" evaluated over a concentration grid reproduces the
population prediction exactly, and a larger cohort would add nothing. The grid
below spans the range of serum concentrations actually observed in the trial,
from below the IC50 up to the highest mean Cmax (2073 ug/mL, part B multiple
60 mg/kg; Table 2).

```{r cohort}
conc_grid <- c(0, exp(seq(log(0.5), log(3000), length.out = 200)))

events <- tibble::tibble(
  id                 = 1L,
  time               = seq_along(conc_grid) - 1,
  evid               = 0L,
  amt                = 0,
  CP_SUTIMLIMAB_UGML = conc_grid
)
```

# Simulation

```{r simulate}
mod <- readModelDb("Bartko_2018_sutimlimab")

sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep   = "CP_SUTIMLIMAB_UGML"
) |>
  as.data.frame()

head(sim[, c("time", "CP_SUTIMLIMAB_UGML", "CPactivity")])
```

# Replicate published figures

## Figure 5 -- CP activity versus serum sutimlimab concentration

```{r figure-5}
# Replicates the "Predicted" curve of Figure 5 of Bartko 2018.
sim |>
  dplyr::filter(CP_SUTIMLIMAB_UGML > 0) |>
  ggplot(aes(CP_SUTIMLIMAB_UGML, CPactivity)) +
  geom_line(linewidth = 1, colour = "#28418c") +
  geom_point(
    data = fig5, aes(conc, digitised),
    inherit.aes = FALSE, shape = 1, size = 2.5, colour = "#c0392b"
  ) +
  geom_hline(yintercept = 94.8 - 90.2, linetype = "dashed") +
  geom_vline(xintercept = 6.2,  linetype = "dotted") +
  geom_vline(xintercept = 15.5, linetype = "dotdash") +
  scale_x_log10(breaks = c(1, 10, 100, 1000, 10000)) +
  coord_cartesian(xlim = c(1, 3000), ylim = c(0, 100)) +
  labs(
    x       = "Sutimlimab serum concentration (ug/mL, log scale)",
    y       = "Classical complement pathway activity (%)",
    title   = "Figure 5 -- sutimlimab concentration versus CP activity",
    caption = paste(
      "Line: packaged model. Open circles: anchors digitised from the published",
      "Figure 5 curve. Dashed horizontal: asymptote E0 - Imax = 4.6%.",
      "Dotted vertical: IC50 = 6.2 ug/mL. Dot-dash vertical: IC90 = 15.5 ug/mL."
    )
  )
```

# PD checks (no PKNCA)

Bartko 2018 characterises sutimlimab PK with model-independent NCA only (Cmax,
tmax, AUC0-inf, AUC0-168 and half-life by cohort in Table 2) and does not fit a
structural population PK model. The packaged model is PD-only -- driven by an
externally supplied `CP_SUTIMLIMAB_UGML` covariate -- so there is no
model-predicted concentration-time profile to integrate and the PKNCA validation
pattern used elsewhere in the package does not apply. The checks below take its
place.

## Closed-form agreement

Every quantity here is deterministic (no etas, residual error fixed to zero), so
the solver output must match the closed form to numerical tolerance.

```{r closed-form}
closed <- e0 - imax * frac(sim$CP_SUTIMLIMAB_UGML)
max_abs_dev <- max(abs(sim$CPactivity - closed))
max_abs_dev
stopifnot(max_abs_dev < 1e-8)
```

## Published anchor points

```{r anchors}
anchor_conc <- c(0, 6.2, 15.5, 100, 1e5)
anchors <- rxode2::rxSolve(
  mod,
  events = tibble::tibble(
    id = 1L, time = seq_along(anchor_conc) - 1, evid = 0L, amt = 0,
    CP_SUTIMLIMAB_UGML = anchor_conc
  ),
  keep = "CP_SUTIMLIMAB_UGML"
) |>
  as.data.frame()

# Fraction of the maximum effect achieved at each anchor.
anchors$frac_of_imax <- (e0 - anchors$CPactivity) / imax

knitr::kable(
  anchors |>
    dplyr::transmute(
      "Sutimlimab (ug/mL)"    = CP_SUTIMLIMAB_UGML,
      "CP activity (%)"       = round(CPactivity, 3),
      "Fraction of Imax"      = round(frac_of_imax, 4),
      "Published anchor"      = c(
        "E0 = 94.8% (Table S3)",
        "IC50 = 6.2 ug/mL -> 50% of Imax (Table S3)",
        "IC90 = 15.5 ug/mL -> 90% of Imax (Abstract, Results)",
        "'concentrations above 100 ug/mL ... near-maximal knockdown' (Results)",
        "asymptote E0 - Imax = 4.6% (Table S3)"
      )
    ),
  caption = "Typical-value evaluation at the concentrations Bartko 2018 anchors explicitly."
)

stopifnot(
  # Zero concentration returns the published baseline exactly.
  abs(anchors$CPactivity[1] - 94.8) < 1e-9,
  # IC50 delivers exactly half of Imax.
  abs(anchors$frac_of_imax[2] - 0.50) < 1e-9,
  # The paper's separately reported IC90 delivers 90% of Imax. This is the
  # strongest single check available: it ties the published IC90 to the
  # published IC50 and Hill coefficient without using any of them twice.
  abs(anchors$frac_of_imax[3] - 0.90) < 0.001,
  # 100 ug/mL is within a quarter of a percentage point of the asymptote,
  # which is what "near-maximal knockdown" means quantitatively.
  anchors$CPactivity[4] - (94.8 - 90.2) < 0.25,
  # The saturating asymptote is E0 - Imax, not E0 * (1 - Imax/100).
  abs(anchors$CPactivity[5] - (94.8 - 90.2)) < 1e-6
)
```

# Comparison against published observations

Bartko 2018 reports no NCA of the PD endpoint, but it does make several
quantitative claims about the depth of CP-activity knockdown at the observed
exposures. Feeding the paper's own mean Cmax values (Table 2) through the model
reproduces them.

```{r cohort-cmax}
cmax_tab <- tibble::tribble(
  ~part,                        ~dose_mgkg, ~cmax_ugml,
  "A, single",                         3,          40,
  "A, single",                        10,         211,
  "A, single",                        30,         602,
  "A, single",                        60,        1464,
  "A, single",                       100,        2036,
  "B, single (first dose)",           30,         653,
  "B, single (first dose)",           60,        1252,
  "B, multiple (4th dose)",           30,         832,
  "B, multiple (4th dose)",           60,        2073
)

cmax_sim <- rxode2::rxSolve(
  mod,
  events = tibble::tibble(
    id = 1L, time = seq_len(nrow(cmax_tab)) - 1, evid = 0L, amt = 0,
    CP_SUTIMLIMAB_UGML = cmax_tab$cmax_ugml
  ),
  keep = "CP_SUTIMLIMAB_UGML"
) |>
  as.data.frame()

cmax_tab <- cmax_tab |>
  dplyr::mutate(
    cp_at_cmax  = cmax_sim$CPactivity,
    suppression = 100 * (e0 - cp_at_cmax) / e0
  )

knitr::kable(
  cmax_tab |>
    dplyr::transmute(
      "Part"                        = part,
      "Dose (mg/kg)"                = dose_mgkg,
      "Published mean Cmax (ug/mL)" = cmax_ugml,
      "Predicted CP activity (%)"   = round(cp_at_cmax, 2),
      "Suppression vs E0 (%)"       = round(suppression, 1)
    ),
  caption = paste(
    "Predicted CP activity at each cohort's published mean Cmax",
    "(Bartko 2018 Table 2)."
  )
)

stopifnot(
  # "complete inhibition, defined by CP activity < 10%, was achieved in all
  # subjects who received a sutimlimab dose of 3 mg/kg or higher" (Discussion).
  all(cmax_tab$cp_at_cmax < 10),
  # "A single infusion of 3, 10, 30, 60, and 100 mg/kg sutimlimab suppressed CP
  # activity by > 90% within 1 hour after the start of the infusion" (Results,
  # Pharmacodynamics).
  all(cmax_tab$suppression > 90)
)
```

Every cohort's mean Cmax sits far above the IC90 of 15.5 ug/mL -- the lowest,
40 ug/mL at 3 mg/kg, is already 2.6-fold above it -- so the model predicts CP
activity between 4.6% and 5.6% at peak in every cohort, i.e. 94-95% suppression.
That is consistent with both published claims, and with the observation that the
0.3 and 1 mg/kg cohorts (whose serum concentrations were below the limit of
quantification throughout) showed "little effect on CP activity".

The model deliberately says nothing about the *duration* of suppression, which
the paper reports as ranging from 8 hours at 3 mg/kg to 14 days at 100 mg/kg.
Duration is a function of the concentration-time profile, and reproducing it
would require a PK model that Bartko 2018 does not provide.

# Assumptions and deviations

- **No structural PK model, by construction.** Bartko 2018 characterises
  sutimlimab PK with model-independent NCA only (Methods, "Pharmacokinetics";
  Table 2) and explicitly reports that elimination is nonlinear below about
  100 ug/mL, consistent with target-mediated disposition: AUC0-168 rose 12.2-fold
  from 3 to 10 mg/kg and 7.7-fold from 10 to 30 mg/kg, but only 2.5- and
  1.6-fold over 30-60 and 60-100 mg/kg, and mean half-life lengthened from 19 to
  132 hours across the 10-100 mg/kg range. A linear compartmental model would not
  capture that, and the paper fits none. The packaged model is therefore PD-only:
  users must supply their own `CP_SUTIMLIMAB_UGML` trajectory (observed
  concentrations, or a PK model from another source). No sutimlimab population PK
  model currently exists in the nlmixr2lib registry.
- **`Imax` enters additively; established by digitising the published Figure 5.**
  Supplementary Table S3 gives `E0 (%)` and `Imax (%)` without the equation. The
  additive reading (asymptote `E0 - Imax` = 4.6%) and the multiplicative reading
  (asymptote `E0 * (1 - Imax/100)` = 9.29%) are both grammatical. The published
  fitted curve spans 94.5% down to a 4.2-4.4% plateau -- a 90.3-percentage-point
  span that reproduces `Imax = 90.2` -- and matches the additive form with an
  RMSE of 0.35 percentage points against 4.32 for the multiplicative form. The
  digitisation is used **only** to choose between two readings of the paper's own
  printed parameters; no parameter value is taken from the figure. See the
  section "Which Imax parameterisation?" above.
- **No interindividual variability and no residual error are reported, so none
  are invented.** Supplementary Table S3 lists exactly four values, each with an
  RSE% (parameter precision, not IIV): E0 94.8 (1.1), Imax 90.2 (1.1), IC50 6.2
  (27.5), H 2.4 (19.9). No omega, no sigma, and no residual-error model appear
  anywhere in the paper or in any of its five supplementary files. `addSd` is
  therefore encoded as `fixed(0)` rather than given an invented magnitude, and
  the model predicts typical values only. A user fitting this model to new data
  should estimate both IIV and residual error from that data. Note that the
  paper's own Figure 5 shows substantial scatter around the fitted line,
  particularly in the dense low-concentration column, so real between-subject and
  residual variability is clearly non-zero -- it is simply not quantified in the
  source.
- **Supplementary Table S3 was retrieved separately from the lead PDF.** The four
  parameter values are not printed in the main text as a set; the main text
  quotes Imax (90.2%), IC50 (6.2 ug/mL), the Hill coefficient (2.4) and the IC90
  (15.5 ug/mL) in prose, but the baseline `E0 = 94.8%` and all four RSE% figures
  appear only in Supplementary Table S3. That file (`CPT-104-655-s003.docx`) was
  obtained from the PMC supplementary-file endpoint for PMC6175298 and its byte
  size and MD5 match the manifest in the Europe PMC full-text XML.
- **Direct effect, no hysteresis.** Bartko 2018 states that exploratory analysis
  found "no delay ... in CP activity" and that individual concentrations and CP
  activities were therefore time-matched (Results, "PK/PD correlations"). The
  packaged model has no effect compartment. The supporting hysteresis analysis
  itself is described as "results retained on file" and is not reproducible from
  the publication.
- **Healthy volunteers only.** The trial's part C, in patients with a
  complement-mediated disorder (cold agglutinin disease, warm autoimmune
  hemolytic anemia, bullous pemphigoid, or antibody-mediated transplant
  rejection), was ongoing at publication and is not reported here. The paper
  cautions that PD may differ in the target populations. The packaged model's
  parameters are healthy-volunteer estimates.
- **Anti-drug antibodies are not modelled.** Confirmed ADA was detected in two
  part A subjects (42 and 28 ng/mL) and in four part B subjects. The paper argues
  the measured ADA concentrations were 500-1000-fold below the drug levels needed
  for a PD effect and reports no distinguishable CP-activity difference, so no
  ADA covariate is included. One part B 30 mg/kg subject with pre-existing ADA
  did show faster CP-activity reversal (Figure S1); that individual behaviour is
  not captured by a typical-value model.
- **Body weight is not a covariate.** Figure 3 shows negative regression slopes
  of AUClast, Cmax, and half-life against body weight, but the authors
  deliberately did **not** perform a formal covariate analysis (fewer than 50
  subjects), and describe those plots as exploratory only. No weight effect is
  encoded. Note also that dosing was already mg/kg, so body size is partly
  accounted for in the dose.
- **Observation variable name `CPactivity` (not `Cc`).**
  `checkModelConventions()` warns that `CPactivity` is not a canonical
  single-output observation name. The convention default `Cc` is reserved for
  drug concentrations; the Bartko 2018 observation is classical complement
  pathway activity expressed as a percentage of the assay's normal reference, not
  a concentration. The paper's own term is retained for source-trace fidelity, no
  rename is performed, and no new canonical is registered. This is the same
  accepted deviation carried by `Weber_1993_remikiren` (observation `APR`), the
  closest structural sibling in the registry.
- **Units of CP activity.** CP activity is a semiquantitative readout of the
  commercial WIESLAB Complement System Classical Pathway enzyme immunoassay
  (Euro Diagnostica, Malmo, Sweden), reported as a percentage of the assay's
  normal reference. Values above 100% are therefore possible and are visible in
  Figure 5; the model's `E0` of 94.8% is the fitted population baseline, close to
  but not identical with the observed group means (97% part A, 94% part B).
