---
title: "Thioguanine (Bayoumy 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Thioguanine (Bayoumy 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Bayoumy_2025_thioguanine"))$reference`
* Description: `r rxode2::rxode(readModelDb("Bayoumy_2025_thioguanine"))$description`
* Article: <https://doi.org/10.1007/s40262-025-01532-1>
* Supplement (ESM 1, final NONMEM control stream): <https://doi.org/10.1007/s40262-025-01532-1>

This is the first published population pharmacokinetic model of thioguanine
(TG) in inflammatory bowel disease. It was built to support model-informed
precision dosing (MIPD): TG is given at a fixed dose in IBD, and therapeutic
drug monitoring of the active metabolites, the 6-thioguanine nucleotides
(6-TGN), is already routine in clinical practice.

Two features of this model are unusual enough to state up front, because both
are easy to mis-transcribe from the printed paper alone (see **Errata** below):

1. **The model runs in days, not hours.** ESM 1 estimates `Ka = 8 /day` and
   `CL/F = 110.405 L/day`; Table 2 of the paper reports the same values per
   hour (0.33 /h, 4.6 L/h).
2. **The apparent volume carries a x5 assay unit-conversion factor applied
   inside `$PK`.** ESM 1 sets `V_FACT = 5` and
   `MU_2 = LOG(THETA(2) * V_FACT) + ...`, so the volume that scales the
   central amount -- and therefore sets `kel = CL/V` -- is
   `270.46 x 5 = 1352.3 L`, not the 270.5 L printed in Table 2. The factor
   converts nmol/L into the historical 6-TGN assay unit, picomole per
   8e8 red blood cells, which the authors equate with picomole per 200 uL of
   blood (Methods 2.3).

Together these give a 6-TGN elimination half-life of about 8.5 days, which is
why the source study sampled troughs out to 8 weeks.

# Population

The model was developed from 131 6-TGN trough measurements in 28 adults with
IBD (16 Crohn's disease, 12 ulcerative colitis) treated with oral thioguanine,
re-analysed from the cohort of Derijks et al. (reference [9] of the source
paper). Six patients were male and 22 female (78.6% female); mean age was 38
years (range 19-70) and median weight 60 kg (IQR 56-74.25). The mean TG dose
was 0.32 mg/kg/day (95% CI 0.29-0.34). Concomitant aminosalicylate (5-ASA)
therapy was taken by 23/28 patients (82%), at a median 5-ASA dose of 3000 mg
(range 1500-3200). TPMT was genotyped in 17/28 patients (61%), all of whom
were `*1/*1` wild type, so a TPMT effect could not be estimated. These are the
baseline characteristics of Table 1 of the source paper.

Trough samples were drawn at days 7, 14, 21, 28 and 56 of daily dosing. Three
further patients contributed a densely sampled 24-h steady-state profile;
because a joint fit destabilised the model and between-occasion variability
could not be estimated from three subjects, those profiles were entered as
separate subjects. Seven of the 131 measurements (5.0%) were removed as
outliers -- three for `|CWRES| > 3` and four for suspected non-adherence.
6-TGN was measured in erythrocytes by reversed-phase HPLC with UV detection
(run-to-run CV 6.6%, LLOQ 30 pmol per 8e8 RBC).

The same information is available programmatically via the model's
`population` metadata (`readModelDb("Bayoumy_2025_thioguanine")()$population`).

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Bayoumy_2025_thioguanine.R` carries an in-file
comment naming its origin. The table below collects them for review. "ESM 1"
is *Supplement 1. Final model code*, the final NONMEM control stream in the
paper's Electronic Supplementary Material.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` | `log(8)`, fixed | ESM 1 `$THETA` 1: `(0.1, 8) FIX ; 1 Ka /day`. Table 2 prints `0.33 (f)` /h = 8 /day. Results 3.2: Ka could not be estimated, so it was fixed at the value best describing the three 24-h profiles. |
| `lcl` | `log(110.405)` | ESM 1 `$THETA` 3: `(0,110.405) ; 3 CL/F L/day`. Table 2 prints 4.6 (= 110.405/24) with RSE 11%. |
| `lvc` | `log(270.46 * 5)` | ESM 1 `$THETA` 2: `(0,270.46) ; 2 V/F`, multiplied by `V_FACT = 5` in `$PK` (`MU_2 = LOG(THETA(2) * V_FACT) + ...`). Table 2 prints the unscaled 270.5 with RSE 15%. |
| `e_wt_cl` | `0.75`, fixed | ESM 1 `$THETA` 6: `(0,0.75) FIX ; 6 WT on CL`. Results 3.3 and the final model equations. |
| `e_wt_vc` | `1.0`, fixed | ESM 1 `$THETA` 5: `(0,1.0) FIX ; 5 WT on V`. Results 3.3 and the final model equations. |
| `e_conmed_amino_cl` | `1.58165` | ESM 1 `$THETA` 7: `(0,1.58165) ; 7 USE of 5-ASA on CL`. Printed as `1.582` in the final model equation and as "58% higher CL" (RSE 15%) in Results 3.3. |
| `etalvc` (omega^2) | `0.271878` | ESM 1 `$OMEGA BLOCK(2)`, first diagonal; ETA(2) is assigned to V in `$PK`. sqrt = 0.521 (52% CV). |
| `etalcl` (omega^2) | `0.150194` | ESM 1 `$OMEGA BLOCK(2)`, second diagonal; ETA(3) is assigned to CL in `$PK`. sqrt = 0.388 (39% CV). |
| `etalvc ~ etalcl` covariance | `0.119927` | ESM 1 `$OMEGA BLOCK(2)` off-diagonal; correlation 0.59. |
| `etalka` (omega^2) | `1`, fixed | ESM 1 `$OMEGA 1 FIX ; BSV Ka`. Results 3.2: BSV on Ka was >250% with large uncertainty, so it was held at 100%. |
| `etapropSd` (omega^2) | `0.244129` | ESM 1 `$OMEGA 0.244129 ; eps` -> ETA(4). sqrt = 0.494 (49% CV). Table 2 "BSV 49%" on the residual-error magnitude. |
| `propSd` | `0.145424` | ESM 1 `$THETA` 4: `(0,0.145424) ; 4 prop error`. Table 2 "Residual error magnitude 14.5%". |
| `CL_i = CL_pop * (WT/70)^0.75 * 1.582^ASA * exp(eta)` | n/a | Final model equation, Results 3.3 (p. 1259); ESM 1 `$PK` `MU_3`. |
| `V_i = V_pop * (WT/70) * exp(eta)` | n/a | Final model equation, Results 3.3 (p. 1259); ESM 1 `$PK` `MU_2`. |
| `d/dt(depot)`, `d/dt(central)` | n/a | One-compartment model with first-order absorption, Results 3.2 and Fig. 1; ESM 1 uses a two-compartment-numbering ADVAN with `S2 = V`. |
| `f(depot) = 1e6/167.19` | n/a | ESM 1 `$PK`: `F1 = 1E6/167.19 ; convert mg dose to nanomole, Mw 6-TG = 167.19 g/mol`. |
| `Cc ~ prop(propSd * exp(etapropSd))` | n/a | ESM 1 `$ERROR`: log-transform-both-sides with `PROP = THETA(4)*EXP(ETA(4))`, `W = PROP`, `Y = LOG(F) + W*EPS(1)`, `$SIGMA 1 FIX`. Results 3.2 states this is "equivalent to a proportional error model". |

# Structural checks

Before simulating a cohort, confirm the covariate model and the unit
conversion are wired up as the control stream specifies. These checks compare
the solved model against its own closed form using the *same* parameter
values, so the difference is pure numerical error and a tight bound is
appropriate.

```{r structural}
mod <- readModelDb("Bayoumy_2025_thioguanine")

MW_TG   <- 167.19          # g/mol, ESM 1 $PK
MG2NMOL <- 1e6 / MW_TG     # mg -> nmol
CL_POP  <- 110.405         # L/day, ESM 1 $THETA 3
V_POP   <- 270.46 * 5      # L, ESM 1 $THETA 2 x V_FACT
E_ASA   <- 1.58165         # ESM 1 $THETA 7

# Typical-value solve over a weight x 5-ASA grid. zeroRe() removes the
# random effects; omega/sigma = NA stops rxSolve reusing a previous draw.
grid <- expand.grid(WT = c(45, 60, 70, 95), CONMED_AMINO = c(0, 1))
grid$id <- seq_len(nrow(grid))

ev_grid <- dplyr::bind_rows(lapply(grid$id, function(i) {
  dplyr::bind_rows(
    data.frame(id = i, time = 0, evid = 1, amt = 20, cmt = "depot"),
    data.frame(id = i, time = c(0, 1, 5), evid = 0, amt = NA_real_, cmt = "central")
  )
})) |>
  dplyr::left_join(grid, by = "id") |>
  dplyr::arrange(id, time, dplyr::desc(evid))

chk <- rxode2::rxSolve(rxode2::zeroRe(mod), ev_grid,
                       omega = NA, sigma = NA,
                       keep = c("WT", "CONMED_AMINO"),
                       returnType = "data.frame") |>
  dplyr::distinct(id, cl, vc, kel, WT, CONMED_AMINO) |>
  dplyr::mutate(
    cl_closed = CL_POP * (WT / 70)^0.75 * E_ASA^CONMED_AMINO,
    vc_closed = V_POP  * (WT / 70)^1.0
  )

stopifnot(
  # Covariate model reproduced exactly (solver round-off only).
  max(abs(chk$cl / chk$cl_closed - 1)) < 1e-8,
  max(abs(chk$vc / chk$vc_closed - 1)) < 1e-8,
  # 5-ASA raises CL/F by exactly the published factor.
  all(abs(chk$cl[chk$CONMED_AMINO == 1] /
          chk$cl[chk$CONMED_AMINO == 0] - E_ASA) < 1e-8)
)

thalf_pop <- log(2) / (CL_POP / V_POP)
knitr::kable(
  chk |>
    dplyr::select(WT, CONMED_AMINO, cl, vc, kel) |>
    dplyr::mutate(`t1/2 (day)` = log(2) / kel) |>
    dplyr::rename("Weight (kg)" = WT, "5-ASA" = CONMED_AMINO,
                  "CL/F (L/day)" = cl, "V/F (L)" = vc, "kel (1/day)" = kel),
  digits = c(0, 0, 1, 1, 5, 2),
  caption = paste0("Typical-value parameters over a weight x 5-ASA grid. ",
                   "The 6-TGN half-life is weight-independent at ",
                   round(thalf_pop, 2), " days because CL/F and V/F ",
                   "scale with different allometric exponents only through ",
                   "their ratio WT^-0.25.")
)
```

The x5 factor is load-bearing: dropping it would shorten the 6-TGN half-life
from `r round(thalf_pop, 1)` days to `r round(log(2) / (CL_POP / 270.46), 1)`
days and put steady state inside the first week, contradicting the source
study's 8-week trough design.

# Virtual cohort

The original data are not public. The cohort below approximates Table 1 of the
source paper: median weight 60 kg and the paper's mean dose of 0.32 mg/kg/day.
Two arms of 200 subjects isolate the 5-ASA covariate; the source cohort was
82% 5-ASA users, but a balanced design is what makes the covariate effect
readable against the paper's two published group means.

```{r cohort}
# set.seed() seeds R's RNG, not rxode2's per-thread simulation streams, so
# the drawn etas differ across machines with different thread counts. Every
# assertion below is written to hold for any cohort this model can produce.
set.seed(20250908)

N_PER_ARM <- 150L
DOSE_MGKG <- 0.32          # mg/kg/day, Table 1 mean
LLOQ      <- 30            # pmol per 8e8 RBC, Methods 2.1

# Dosing duration for the NCA arm. The closed-form gates below are asserted
# with a tight `all()` bound, which is only legitimate if EVERY subject is
# genuinely at steady state -- otherwise the residual approach to steady
# state is a per-subject physical mechanism and the extreme of a random
# cohort is not reproducible across rxode2 builds. kel = cl/vc has a
# log-scale SD of sqrt(0.150194 + 0.271878 - 2*0.119927) = 0.427, so a
# 4-SD-slow subject has a half-life near 47 days. Dosing for 500 days gives
# even that subject more than ten half-lives.
DAYS_DOSE <- 500L
stopifnot(
  log(2) / (CL_POP / V_POP * exp(-4 * sqrt(0.150194 + 0.271878 - 2 * 0.119927))) * 10 <
    DAYS_DOSE
)

make_arm <- function(n, asa, arm, id_offset = 0L) {
  tibble::tibble(
    id  = id_offset + seq_len(n),
    # Weight: lognormal with the published median of 60 kg. The paper gives
    # only median and IQR, so the spread (sdlog = 0.20) is an assumption --
    # see "Assumptions and deviations".
    WT  = pmin(pmax(stats::rlnorm(n, log(60), 0.20), 40), 120),
    CONMED_AMINO = asa,
    arm = arm
  )
}

cohort <- dplyr::bind_rows(
  make_arm(N_PER_ARM, 0, "No 5-ASA",          id_offset = 0L),
  make_arm(N_PER_ARM, 1, "Concomitant 5-ASA", id_offset = N_PER_ARM)
)

# Observation grid: pre-dose troughs (offset just before the dose so the
# record is unambiguously a trough) -- daily over the source study's 8-week
# window, then weekly out to the NCA interval -- plus a dense grid over the
# final steady-state dosing interval and a washout for the half-life.
t_trough_daily  <- seq_len(56) - 1e-3
t_trough_weekly <- seq(63, DAYS_DOSE - 1, by = 7) - 1e-3
t_ss            <- seq(DAYS_DOSE, DAYS_DOSE + 1, by = 0.04)
t_washout       <- seq(DAYS_DOSE + 2, DAYS_DOSE + 70, by = 4)
t_obs <- sort(unique(c(0, t_trough_daily, t_trough_weekly, t_ss, t_washout)))

events <- dplyr::bind_rows(
  # One dose record per subject, expanded by rxode2 via ii/addl into daily
  # oral doses on days 0 .. DAYS_DOSE.
  cohort |>
    dplyr::mutate(time = 0, evid = 1, amt = DOSE_MGKG * WT, cmt = "depot",
                  ii = 1, addl = DAYS_DOSE),
  cohort |>
    tidyr::expand_grid(time = t_obs) |>
    dplyr::mutate(evid = 0, amt = NA_real_, cmt = "central",
                  ii = 0, addl = 0)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(
  nrow(dplyr::distinct(cohort, id)) == 2L * N_PER_ARM,
  N_PER_ARM <= 200L
)
```

# Simulation

```{r simulate}
simres <- rxode2::rxSolve(
  mod, events,
  keep = c("WT", "CONMED_AMINO", "arm"),
  returnType = "data.frame"
) |>
  dplyr::filter(!is.na(Cc))

# `Cc` is the individual prediction; `sim` additionally carries the
# per-subject proportional residual error. Apply the assay LLOQ to `sim` the
# way the laboratory would: values below the limit are reported as LLOQ/2.
# Without this the large residual error occasionally drives a simulated
# concentration negative, which no assay can return.
simres <- simres |>
  dplyr::mutate(dv = pmax(sim, LLOQ / 2))

stopifnot(
  !anyNA(simres$Cc),
  all(simres$Cc >= 0),
  all(simres$dv > 0)
)
```

# Replicate published figures

The source paper's Fig. 2 is a visual predictive check of 6-TGN trough
concentration against time over the 8-week trough cohort, with the median and
the 5th-95th percentile prediction interval. The panel below reproduces that
shape from the packaged model over the same 8-week window, split by 5-ASA
status so the covariate effect is visible.

```{r figure-2}
# Replicates Figure 2 of Bayoumy 2025: VPC of 6-TGN trough vs time.
simres |>
  dplyr::filter(time %in% t_trough_daily) |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(
    Q05 = stats::quantile(dv, 0.05),
    Q50 = stats::quantile(dv, 0.50),
    Q95 = stats::quantile(dv, 0.95),
    .groups = "drop"
  ) |>
  ggplot2::ggplot(ggplot2::aes(time, Q50)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::facet_wrap(~arm) +
  ggplot2::labs(
    x = "Time (days)",
    y = "6-TGN (pmol per 8e8 RBC)",
    title = "Figure 2 - simulated 6-TGN trough accumulation over 8 weeks",
    caption = paste("Replicates Figure 2 of Bayoumy 2025 (VPC).",
                    "Line = median, band = 5th-95th percentile.")
  )
```

Accumulation to steady state takes roughly six weeks, consistent with the
`r round(thalf_pop, 1)`-day 6-TGN half-life and with the source study's choice
to sample out to day 56.

# PKNCA validation

NCA is run over the final steady-state dosing interval (day
`r DAYS_DOSE` to `r DAYS_DOSE + 1`) for exposure parameters, and over the
washout window for the terminal half-life.

```{r pknca}
# Only `!is.na(Cc)` -- do NOT filter on time or concentration, which would
# drop the time-zero anchor PKNCA needs.
sim_nca <- simres |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

# Guarantee a time-zero record per subject; pre-dose Cc = 0 is correct for an
# extravascular dose into an empty compartment.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, arm, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

# The rxSolve event table carries one ii/addl dose record per subject;
# PKNCA needs the dose that opens the steady-state interval, so expand just
# that one. No dose falls in the washout interval, which is correct.
dose_df <- cohort |>
  dplyr::mutate(time = DAYS_DOSE, amt = DOSE_MGKG * WT) |>
  dplyr::select(id, time, amt, arm)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)

intervals <- data.frame(
  start      = c(DAYS_DOSE, DAYS_DOSE + 1),
  end        = c(DAYS_DOSE + 1, DAYS_DOSE + 70),
  cmax       = c(TRUE,  FALSE),
  tmax       = c(TRUE,  FALSE),
  auclast    = c(TRUE,  FALSE),
  cav        = c(TRUE,  FALSE),
  half.life  = c(FALSE, TRUE)
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))

# Split by interval before reshaping: `tmax` is returned for BOTH intervals
# (and the washout interval also returns the lambda.z diagnostics), so a
# single pivot_wider would collide and silently produce list-columns.
nca_long <- as.data.frame(nca_res) |>
  dplyr::select(arm, id, start, PPTESTCD, PPORRES)

nca_wide <- dplyr::left_join(
  nca_long |>
    dplyr::filter(start == DAYS_DOSE,
                  PPTESTCD %in% c("cmax", "tmax", "auclast", "cav")) |>
    dplyr::select(-start) |>
    tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES),
  nca_long |>
    dplyr::filter(start == DAYS_DOSE + 1, PPTESTCD == "half.life") |>
    dplyr::select(arm, id, half.life = PPORRES),
  by = c("arm", "id")
)
stopifnot(nrow(nca_wide) == 2L * N_PER_ARM,
          all(vapply(nca_wide, is.atomic, logical(1))))

knitr::kable(
  nca_wide |>
    dplyr::group_by(arm) |>
    dplyr::summarise(
      cmax = stats::median(cmax), tmax = stats::median(tmax),
      auclast = stats::median(auclast), cav = stats::median(cav),
      half.life = stats::median(half.life), .groups = "drop"
    ) |>
    dplyr::rename("Arm" = arm,
                  "Cmax (pmol/8e8 RBC)" = cmax,
                  "Tmax (day)" = tmax,
                  "AUCtau (pmol*day/8e8 RBC)" = auclast,
                  "Cav (pmol/8e8 RBC)" = cav,
                  "t1/2 (day)" = half.life),
  digits = 1,
  caption = paste0("Median simulated steady-state NCA over the day ",
                   DAYS_DOSE, "-", DAYS_DOSE + 1,
                   " dosing interval, with the terminal half-life from the ",
                   "post-dosing washout.")
)
```

## Closed-form gates on the NCA output

Two identities must hold if the unit conversion and the disposition model are
encoded correctly. Both compare the NCA output against each subject's *own*
solved parameters, so they are numerical-accuracy checks and are asserted
tightly.

```{r nca-gates}
subj <- simres |>
  dplyr::distinct(id, arm, WT, cl, vc, kel) |>
  dplyr::left_join(nca_wide, by = c("id", "arm")) |>
  dplyr::mutate(
    dose_nmol = DOSE_MGKG * WT * MG2NMOL,
    # Steady-state mass balance: Cav * CL * tau == dose delivered per interval.
    mass_bal  = cav * cl * 1 / dose_nmol,
    # Terminal half-life must equal log(2)/kel for each subject.
    thalf_rel = half.life / (log(2) / kel)
  )

stopifnot(
  !anyNA(subj$cav), !anyNA(subj$half.life),
  # Mass balance: within 1% of unity for every subject. The small systematic
  # shortfall is the residual approach to steady state at day 90 plus
  # trapezoidal error on the absorption peak.
  all(abs(subj$mass_bal - 1) < 0.01),
  # Half-life recovered from the washout to within 1% for every subject.
  all(abs(subj$thalf_rel - 1) < 0.01)
)

knitr::kable(
  subj |>
    dplyr::group_by(arm) |>
    dplyr::summarise(
      n = dplyr::n(),
      mass_bal_med = stats::median(mass_bal),
      mass_bal_max = max(abs(mass_bal - 1)),
      thalf_med    = stats::median(thalf_rel),
      thalf_max    = max(abs(thalf_rel - 1)),
      .groups = "drop"
    ) |>
    dplyr::rename("Arm" = arm, "N" = n,
                  "Cav*CL/Dose (median)" = mass_bal_med,
                  "Cav*CL/Dose (max |dev|)" = mass_bal_max,
                  "t1/2 NCA / closed form (median)" = thalf_med,
                  "t1/2 ratio (max |dev|)" = thalf_max),
  digits = 5,
  caption = paste("Closed-form gates. A value of 1 is exact; the maximum",
                  "absolute deviation across all subjects is shown.")
)
```

# Comparison against published values

The source paper reports no NCA parameters -- the analysis dataset is almost
entirely trough concentrations, so Cmax, Tmax and AUC were never computed. It
does, however, publish one directly comparable exposure statistic. In
Limitations 4.2, to corroborate the model-based 5-ASA effect, the authors
report the mean steady-state trough at 4 weeks in each group: **805** with
concomitant 5-ASA and **1070** without (a significant difference, p < 0.05).

Those means are the reference below, compared against the simulated
week-4 (day 28) trough. Because the 6-TGN half-life is
`r round(thalf_pop, 1)` days against a 1-day dosing interval, peak-to-trough
fluctuation is only about
`r round(100 * (exp(CL_POP / V_POP) - 1), 1)`%, so the day-28 trough is a
faithful stand-in for average exposure.

```{r nca_compare}
sim_trough_wk4 <- simres |>
  dplyr::filter(abs(time - (28 - 1e-3)) < 1e-9) |>
  dplyr::select(arm, id, ctrough = Cc)

published <- tibble::tribble(
  ~arm,                 ~ctrough,
  "No 5-ASA",              1070,
  "Concomitant 5-ASA",      805
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = sim_trough_wk4 |> dplyr::select(arm, ctrough),
  reference     = published,
  by            = "arm",
  units         = c(ctrough = "pmol per 8e8 RBC"),
  tolerance_pct = 25
)

knitr::kable(
  cmp,
  caption = paste("Simulated vs published mean 4-week 6-TGN trough",
                  "(Bayoumy 2025, Limitations 4.2).",
                  "* differs from the reference by >25%."),
  align = c("l", "l", "r", "r", "r")
)
```

```{r nca_compare_gate}
gate <- sim_trough_wk4 |>
  dplyr::group_by(arm) |>
  dplyr::summarise(med = stats::median(ctrough), .groups = "drop") |>
  dplyr::left_join(published, by = "arm") |>
  dplyr::mutate(pct_diff = 100 * (med - ctrough) / ctrough)

stopifnot(
  # Structural: a mis-transcribed clearance, dose, time base or unit factor
  # moves these by a factor, not by tens of percent.
  all(abs(gate$pct_diff) < 30),
  # The published direction of the 5-ASA effect must be reproduced: higher
  # clearance on 5-ASA means a lower trough.
  gate$med[gate$arm == "Concomitant 5-ASA"] < gate$med[gate$arm == "No 5-ASA"]
)

knitr::kable(
  gate |>
    dplyr::rename("Arm" = arm,
                  "Simulated median trough" = med,
                  "Published mean trough" = ctrough,
                  "% difference" = pct_diff),
  digits = 1,
  caption = "Week-4 trough: simulated median vs the published group mean."
)
```

The no-5-ASA arm lands within a few percent of the published 1070. The
5-ASA arm simulates lower than the published 805, which is expected and not a
transcription error: 1070/805 = 1.33 is the *observed* group-mean ratio in an
unbalanced 19-vs-4 split, whereas the model's fitted covariate effect is
1.582 on clearance. The authors themselves present the group means only as
corroboration of the direction of the effect, not as a calibration target,
and they note that 5-ASA dose could not be related to clearance at all.

# Assumptions and deviations

## Errata and conflicts in the source

- **Table 2's `Unit` column is transposed between its two disposition rows.**
  It prints "CL/F ... L" and "V/F ... L/h". CL/F is a clearance and V/F is a
  volume; ESM 1 labels them `; 3 CL/F L/day` and `; 2 V/F`. The control
  stream was followed.
- **Table 2's `BSV` column disagrees with ESM 1 about which parameter carries
  which variance.** ESM 1 assigns `ETA(2)` to V and `ETA(3)` to CL in `$PK`,
  and its `$OMEGA BLOCK(2)` declares `0.271878 ; BSV V` then
  `0.119927 0.150194 ; BSV CL` -- that is, omega_V = 52% and omega_CL = 39%.
  Table 2, and Sections 3.4 and 4 which read off it, assign 52% to CL/F and
  39% to V/F, the opposite way round. **This model follows the control
  stream**, because it is the final model code, its ETA ordering and its own
  inline comments agree with each other, and the adjacent `Unit` column of the
  same two Table 2 rows is demonstrably transposed as well. The two variance
  magnitudes themselves are not in dispute -- only which parameter each
  belongs to. A reviewer who prefers Table 2 need only swap the two diagonals
  of the `etalvc + etalcl` block. Note that this choice does not affect any
  gate in this vignette: all of them are conditioned on each subject's own
  solved `cl` and `vc`.
- **Section 3.4 mislabels the RSE column as residual error** ("Residual error
  was estimated to be 15% and 11% for V/F and CL/F"); Table 2 lists those
  values under `RSE`, and the residual error magnitude is the separate 14.5%
  row.
- **The trough concentrations in Limitations 4.2 are printed as "805 mg/L"
  and "1070 mg/L".** 6-TGN is reported in pmol per 8e8 RBC throughout the
  rest of the paper; mg/L is not a possible unit for this assay at this
  magnitude. The values are used here in the assay's own units.
- **Section 3.4 attributes the final parameters to "the base model"** while
  Table 2 is captioned as the model's PopPK parameters and the Discussion
  treats them as final. ESM 1 is titled "Final model code" and contains
  exactly these values, so they are treated as the final estimates.
- **`n_observations = 131` predates outlier removal.** The paper says 7
  concentrations (5.0%) were removed; 7/131 is 5.3%, so the 5.0% figure is
  presumably computed against a slightly larger pre-cleaning set that is not
  reported.

## Modelling assumptions made here

- **Weight distribution.** The paper reports only the median (60 kg) and IQR
  (56-74.25). The cohort uses a lognormal with that median and `sdlog = 0.20`,
  truncated to 40-120 kg. The published IQR is markedly asymmetric about its
  median and no single lognormal reproduces both quartiles; the spread is
  therefore an assumption. No gate in this vignette depends on it -- the
  closed-form gates are per-subject and the trough comparison uses each
  subject's own weight-scaled clearance.
- **5-ASA arm balance.** Simulated 200/200 rather than the cohort's 23/5, to
  make the covariate effect readable. The published group means are compared
  arm-by-arm, so the imbalance in the source does not enter.
- **Dose.** `0.32 mg/kg/day`, the cohort mean of Table 1, given as a
  continuous per-subject amount. In practice TG is given as a fixed dose in
  IBD; a fixed 20 mg/day would shift both arms by the ratio of actual to
  weight-proportional dose without changing any structural conclusion.
- **Steady state.** Dosing runs to day 500 rather than the source study's day
  56. This is not cosmetic: the closed-form mass-balance gate is asserted with
  a tight `all()` bound across every subject, and that is only legitimate once
  the residual approach to steady state has been eliminated. At day 56 a
  subject drawing a slow `kel` would still be short of steady state, and how
  slow the slowest subject is depends on the random draw -- which rxode2
  partitions per solver thread and therefore varies with the machine's thread
  count. Dosing to day 500 gives even a 4-SD-slow subject more than ten
  half-lives, so the residual is numerical rather than physical. The 8-week
  accumulation figure and the week-4 trough comparison still use the source
  study's own window.
- **LLOQ handling.** Simulated observations carrying residual error are
  floored at LLOQ/2 = 15 pmol per 8e8 RBC (Methods 2.1 gives an LLOQ of 30).
  The residual-error model is proportional with a per-subject magnitude whose
  BSV is 49%, so without the floor a small number of draws in the
  highest-error subjects would be negative. Only the VPC figure uses these
  values; every gate uses `Cc`.
- **Pre-existing 6-TGN not modelled.** ESM 1 `$PK` carries a data-handling
  branch, `IF(A_0FLG.EQ.1) A_0(2) = CONC0 * V`, initialising the central
  compartment for patients who switched from azathioprine or mercaptopurine
  without a full washout. That is a property of the analysis dataset, not of
  the structural model, so it is not encoded; simulations here start from an
  empty compartment.
- **Between-occasion variability not modelled.** ESM 1 contains a commented-out
  IOV block. The paper states IOV could not be estimated from the three
  densely sampled subjects and was not retained in the final model.
- **No non-paper-derived parameter values.** Every `ini()` entry comes from
  the paper's Table 2, its final model equations, or ESM 1. Nothing was
  digitised from a figure, obtained by correspondence, or carried from an
  upstream publication.
