---
title: "Cloxacillin (Beijer 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cloxacillin (Beijer 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Beijer_2026_cloxacillin"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1093/jac/dkag116>
* Supplement (Figures S1-S4, Tables S1-S2, appendices, open access via EuropePMC):
  <https://www.ebi.ac.uk/europepmc/webservices/rest/PMC13049546/supplementaryFiles>

Every parameter value in this model comes from **Table S2 of the Supplementary
Material**; the main article prints no parameter estimates at all. The
protein-binding equation is Supplementary Material Eq. 1 and the structural
schematic that settles which pool the ODE states hold is Figure S1.

# Population

Two hundred adults undergoing primary elective total hip arthroplasty (95, 47%)
or total knee arthroplasty (105, 53%) at two Swedish centres between 2022 and
2024 (Table 1). Median age 73 years (IQR 65-78, range 36-90), median body weight
83 kg (IQR 73-95, range 53-185), 105/200 (53%) female. Renal function was
generally well preserved: median relative eGFR 72 mL/min/1.73 m^2 (IQR 61-84,
range 18-142), median plasma creatinine 70 umol/L. Median plasma albumin 34 g/L.
ASA class I 15%, II 42%, III 43%.

All patients received 2 g of intravenous cloxacillin 30-45 min before incision,
repeated 2 h and 6 h after the start of the first dose, per Swedish national
guidelines. Adherence was poor: only 42/200 (21%) preoperative doses complied
with the guideline in full, infusion durations were under 20 min in 90 (45%) of
patients, and 63 (32%) received the dose more than 45 min before incision.

Total and unbound plasma cloxacillin were both assayed by reversed-phase
HPLC-MS/MS, with the unbound fraction separated by ultrafiltration through a
10 kDa membrane. 496 samples were evaluable, median 3 per patient. Observed
median plasma protein binding was 91% (range 69-98%), and 98/200 patients (49%)
fell below 90% binding in at least one sample -- lower and far more variable
than the 92-94% that cloxacillin monographs report from healthy volunteers.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Beijer_2026_cloxacillin")()$population`).

# What the ODE states hold

This is the one structural point that has to be right before anything else in
the vignette means what it says, so it is worth stating explicitly.

Supplementary Figure S1 draws two solid-outlined compartments, `V1` (plasma) and
`V2` (extravascular space), each holding an **unbound** concentration, with `CL`
leaving `V1` and `Q` connecting the two. The bound plasma pool hangs off `V1` as
a dashed box coupled by `Kd` and `Bmax` -- an equilibrium, not a mass-transfer
arm -- and the bound tissue pool is greyed out entirely because the study never
measured it.

So the disposition is parameterised end to end on unbound cloxacillin. `central`
and `peripheral1` both hold unbound drug, `central / vc` is the unbound plasma
concentration, and the total concentration is recovered from it algebraically:

$$C_{tot} = C_u + \frac{B_{max} \cdot C_u}{K_d + C_u} \qquad \text{(Supplementary Material Eq. 1)}$$

Two consequences follow, and both are load-bearing for any reuse of this model:

1. `central / vc` is an **unbound** concentration even though the library
   convention names it `Cc`. The `label()` on `lvc` and the `Cc` comment in the
   model body both record this. `Ctot` is the second output.
2. The binding term is an **output transformation, not a reservoir**. No mass
   moves into it, so `Ctot` does not feed back on the disposition, and the
   administered amount enters `central` unchanged with no bioavailability term.
   The mass-balance gate below confirms that reading: unbound clearance times
   unbound AUC recovers the whole administered dose.

# Source trace

The per-parameter origin is also recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Beijer_2026_cloxacillin.R`.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Model structure (2-cmt on unbound drug; `CL`, `Q` on the unbound pool) | n/a | Figure S1; Results, "A two-compartment model with linear elimination and nonlinear protein binding" |
| `Ctot = Cu + Bmax * Cu / (Kd + Cu)` | n/a | Supplementary Material Eq. 1, "Protein binding" section |
| `CL = theta_CL * (BW/84)^beta1 * (eGFR/67)^beta2 * exp(eta_CL)` | n/a | Table S2, CL row (equation printed in the row header) |
| `lcl` (theta_CL) | 76.4 L/h | Table S2 (RSE 4.4%; bootstrap median 74.3, 95% CI 61.4-83.3) |
| `e_wt_cl` (beta1) | 0.67 | Table S2 (RSE 22%; bootstrap median 0.71, 95% CI 0.39-1.07) |
| `e_crcl_cl` (beta2) | 0.51 | Table S2 (RSE 18%; bootstrap median 0.53, 95% CI 0.33-0.75) |
| WT centring constant | 84 kg | Table S2 CL row `(BW/84)`; corroborated by the Figure 1 caption typical patient |
| CRCL centring constant | 67 mL/min/1.73 m^2 | Table S2 CL row `(eGFR/67)`; corroborated by the Figure 1 caption typical patient |
| `lvc` (theta_V1) | 57.1 L | Table S2 (RSE 7.7%; bootstrap median 56.5, 95% CI 43.8-70.8) |
| `lq` (theta_Q) | 57.2 L/h | Table S2 (RSE 12%; bootstrap median 56.1, 95% CI 42.6-96.3) |
| `lvp` (theta_V2) | 69.0 L | Table S2 (RSE 10%; bootstrap median 72.3, 95% CI 53.4-153) |
| `lbmax` (theta_Bmax) | 559 mg/L | Table S2 (RSE 1.6%; bootstrap median 564, 95% CI 511-625) |
| `lkd` (theta_Kd) | 47.8 mg/L | Table S2 (RSE 2.6%; bootstrap median 48.7, 95% CI 43.0-55.0) |
| `etalcl` variance | 0.2025 | Table S2 `omega_CL` = 0.45 (CV 47%, RSE 7.1%); variance = 0.45^2 |
| `etalvc` variance | 0.4096 | Table S2 `omega_V1` = 0.64 (CV 72%, RSE 9.2%); variance = 0.64^2 |
| `etalcl`-`etalvc` covariance | 0.19584 | Table S2 `rho_CL~V1` = 0.68 (RSE 9.8%); covariance = 0.68 x 0.45 x 0.64, per Supplementary Material "Correlation between parameters" |
| `etalq` variance | 0.6241 | Table S2 `omega_Q` = 0.79 (CV 94%, RSE 13%); variance = 0.79^2 |
| `etalbmax` variance | 0.0025 | Table S2 `omega_Bmax` = 0.05 (CV 4.7%, RSE 25%); variance = 0.05^2 |
| `expSd` (unbound) | 0.24 | Table S2 `a(unbound)` (RSE 4.4%); error model `log(Y) = log(f) + a * eps` |
| `expSd_Ctot` (total) | 0.19 | Table S2 `a(total)` (RSE 4.6%) |

## Omega scale

Table S2 reports each `omega` alongside a CV% in the same cell, which settles
the scale without ambiguity: Monolix's `omega` is the standard deviation of a
log-normally distributed random effect, so the CV is
`sqrt(exp(omega^2) - 1)`. `ini()` takes the variance, so each value is squared.

```{r omega-scale}
omega <- c(CL = 0.45, V1 = 0.64, Q = 0.79, Bmax = 0.05)
printed_cv <- c(CL = 47, V1 = 72, Q = 94, Bmax = 4.7)
derived_cv <- 100 * sqrt(exp(omega^2) - 1)

data.frame(
  Parameter = names(omega),
  omega = omega,
  `Printed CV (%)` = printed_cv,
  `sqrt(exp(omega^2) - 1) (%)` = round(derived_cv, 1),
  check.names = FALSE
) |>
  knitr::kable(row.names = FALSE, caption = "Table S2's CV column reproduced from omega on the log-SD scale.")

# The log-SD reading reproduces every printed CV to within rounding of the
# two-decimal omega. An additive-eta reading would give 100 * omega / theta,
# which for CL is 100 * 0.45 / 76.4 = 0.6% -- nowhere near the printed 47%.
stopifnot(max(abs(derived_cv - printed_cv)) < 1)
```

# Figure 1 -- typical-patient profiles

The Figure 1 caption names the typical patient exactly: "a body weight of 84 kg
and eGFR 67 mL/min/1.73 m2 receiving a 20 min infusion of 2 g cloxacillin".
Those are the two covariate centring constants, so the typical patient is the
model's own reference subject and `zeroRe()` reproduces the figure's
superimposed curves directly.

```{r typical-events}
mod  <- readModelDb("Beijer_2026_cloxacillin")
modT <- rxode2::zeroRe(mod)

# Route A event table (cmt = ODE state + explicit dvid). This model declares
# two endpoints (Cc and Ctot), so every observation row must name which one it
# belongs to; `dvid = 1L` selects Cc and rxSolve returns Ctot alongside it.
# `cmt` stays on the ODE state `central` -- naming an observable there would
# inject a compartment slot and renumber the states.
ev_typical <- function(dose_times, amt, dur, obs_times, WT, CRCL, id = 1L) {
  dose <- data.frame(time = dose_times, amt = amt, dur = dur, rate = NA_real_,
                     cmt = "central", evid = 1L, dvid = NA_integer_)
  obs  <- data.frame(time = obs_times, amt = NA_real_, dur = NA_real_, rate = NA_real_,
                     cmt = "central", evid = 0L, dvid = 1L)
  out <- rbind(dose, obs)
  out <- out[order(out$time, -out$evid), ]
  out$id <- id
  out$WT <- WT
  out$CRCL <- CRCL
  out
}

# The guideline regimen: 2 g at 0, 2 and 6 h, each over 20 min.
d_fig1 <- ev_typical(
  dose_times = c(0, 2, 6), amt = 2000, dur = 20 / 60,
  obs_times  = seq(0, 10, by = 1 / 240),
  WT = 84, CRCL = 67
)
sim_fig1 <- rxode2::rxSolve(modT, d_fig1, returnType = "data.frame")
stopifnot(all(sim_fig1$Cc >= 0), all(sim_fig1$Ctot >= 0))
```

```{r figure-1, echo = FALSE}
sim_fig1 |>
  select(time, Unbound = Cc, Total = Ctot) |>
  pivot_longer(-time, names_to = "Analyte", values_to = "conc") |>
  ggplot(aes(time, conc, colour = Analyte)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = c(1, 2), linetype = "dashed", colour = "grey40") +
  scale_y_log10() +
  scale_colour_manual(values = c(Unbound = "#c0392b", Total = "#2471a3")) +
  labs(
    x = "Time after first dose (h)", y = "Cloxacillin concentration (mg/L)",
    title = "Figure 1 - typical patient (84 kg, eGFR 67), 2 g over 20 min at 0, 2 and 6 h",
    caption = paste(
      "Replicates the superimposed model curves of Figure 1 of Beijer 2026.",
      "Dashed lines are the 1 and 2 mg/L unbound MIC targets."
    )
  )
```

```{r figure-1-gates}
at <- function(tt, col = "Cc") sim_fig1[[col]][which.min(abs(sim_fig1$time - tt))]

# These are deterministic typical-value predictions, not cohort statistics, so
# the bounds are tight on purpose: a mis-transcribed clearance, volume, dose or
# centring constant moves each of them by tens of percent.
gates <- tibble::tribble(
  ~Quantity,                                        ~Published, ~Model,        ~Source,
  "Unbound Cc at 2 h after first dose (mg/L)",             3.2, at(2),         "Results: model-predicted median 3.2 mg/L (3.1 excluding second-dose patients)",
  "Unbound Cc at 0.75 h, ~start of surgery (mg/L)",       10.0, at(0.75),      "Results: observed median 10 mg/L at start of surgery",
  "Plasma protein binding at 0.75 h (%)",                 91.0, 100 * (1 - at(0.75) / at(0.75, "Ctot")), "Results: observed median protein binding 91%"
) |>
  mutate(`Difference (%)` = round(100 * (Model - Published) / Published, 1),
         Model = round(Model, 2))

gates |>
  select(Quantity, Published, Model, `Difference (%)`, Source) |>
  knitr::kable(caption = "Typical-patient predictions against the values Beijer 2026 reports.")

stopifnot(
  # The 2 h concentration is the paper's headline number and the model
  # reproduces it to two significant figures.
  abs(at(2) - 3.2) < 0.15,
  # The start-of-surgery and protein-binding checks compare a typical-value
  # prediction against an observed cohort MEDIAN, so they get a wider band --
  # the sampling time was dictated by surgical events, not by the clock.
  abs(at(0.75) - 10) < 2,
  abs(100 * (1 - at(0.75) / at(0.75, "Ctot")) - 91) < 3
)
```

# Figure 2 -- concentration-dependent unbound fraction

Figure 2 plots the observed unbound fraction against unbound concentration and
overlays the curve that the final `Bmax` and `Kd` estimates imply, with
reference lines at the 92% and 94% linear protein binding that cloxacillin
monographs quote. Inverting Supplementary Material Eq. 1 gives that curve in
closed form, with no simulation needed.

```{r figure-2}
bmax <- 559
kd   <- 47.8
fu_of_cu <- function(cu) cu / (cu + bmax * cu / (kd + cu))

cu_grid <- 10^seq(log10(0.01), log10(60), length.out = 400)
fu_curve <- data.frame(cu = cu_grid, fu = 100 * fu_of_cu(cu_grid))

ggplot(fu_curve, aes(cu, fu)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 100 - c(92, 94), linetype = "dashed", colour = "#c0392b") +
  scale_x_log10() +
  labs(
    x = "Unbound cloxacillin concentration (mg/L)", y = "Unbound fraction (%)",
    title = "Figure 2 - unbound fraction rises with concentration",
    caption = paste(
      "Replicates the non-linear regression curve of Figure 2 of Beijer 2026 from",
      "Bmax = 559 mg/L and Kd = 47.8 mg/L. Dashed lines are the 92% and 94%",
      "linear-binding references (8% and 6% unbound)."
    )
  )
```

```{r figure-2-gates}
# Deterministic algebra, so these are exact checks of the binding equation.
fu_range <- 100 * fu_of_cu(c(2.1, 55))    # observed unbound range, start of surgery
cat(sprintf("Unbound fraction across the observed 2.1-55 mg/L range: %.1f%% to %.1f%%\n",
            fu_range[1], fu_range[2]))
cat(sprintf("Protein binding across that range: %.1f%% to %.1f%%   [observed range 69-98%%]\n",
            100 - fu_range[1], 100 - fu_range[2]))
cat(sprintf("Protein binding at the observed median unbound of 10 mg/L: %.1f%%   [observed median 91%%]\n",
            100 - 100 * fu_of_cu(10)))

stopifnot(
  # The whole point of the non-linear term: unbound fraction is strictly
  # increasing in concentration. Deterministic, so strict monotonicity is safe.
  all(diff(fu_curve$fu) > 0),
  # The model's binding at the observed median unbound concentration matches the
  # observed median binding.
  abs((100 - 100 * fu_of_cu(10)) - 91) < 2,
  # The curve sits inside the observed 69-98% binding envelope over the observed
  # concentration range, and crosses the monograph 92-94% band only at the low
  # end -- which is the paper's point about monograph values being too high.
  all(100 - fu_range > 69), all(100 - fu_range < 98)
)
```

# Virtual cohort

The original patient-level data are not public. The paper's own Monte Carlo
simulations "drew random covariate values for the simulated patients from
distributions that were assumed to be identical to the covariate distributions
in our study population", so the cohort below does the same: log-normal body
weight and relative eGFR matched to the Table 1 median and IQR, truncated to
the Table 1 observed range.

```{r cohort}
# set.seed() seeds R's RNG for the covariate draws below. It does NOT seed
# rxode2's eta sampler, whose streams are partitioned per solver thread, so the
# simulated etas differ between this machine and CI. Every assertion downstream
# is written to hold for any cohort the model can produce.
set.seed(20260913)

n_per_arm <- 200L   # the 200-per-arm cap; the paper simulated 1000

# Fit a log-normal to a median and an IQR: meanlog = log(median),
# sdlog = log(q75 / q25) / (2 * qnorm(0.75)).
lnorm_from_iqr <- function(med, q25, q75) {
  c(meanlog = log(med), sdlog = log(q75 / q25) / (2 * stats::qnorm(0.75)))
}
wt_par   <- lnorm_from_iqr(83, 73, 95)    # Table 1: median 83, IQR 73-95, range 53-185
crcl_par <- lnorm_from_iqr(72, 61, 84)    # Table 1: median 72, IQR 61-84, range 18-142

rtrunc_lnorm <- function(n, par, lo, hi) {
  x <- stats::rlnorm(n, par[["meanlog"]], par[["sdlog"]])
  pmin(pmax(x, lo), hi)
}

subjects <- tibble::tibble(
  subj = seq_len(n_per_arm),
  WT   = rtrunc_lnorm(n_per_arm, wt_par,   53, 185),
  CRCL = rtrunc_lnorm(n_per_arm, crcl_par, 18, 142)
)

tibble::tibble(
  Covariate = c("Body weight (kg)", "Relative eGFR (mL/min/1.73 m^2)"),
  `Cohort median` = round(c(median(subjects$WT), median(subjects$CRCL)), 1),
  `Cohort IQR` = c(
    paste(round(stats::quantile(subjects$WT, c(.25, .75)), 0), collapse = "-"),
    paste(round(stats::quantile(subjects$CRCL, c(.25, .75)), 0), collapse = "-")
  ),
  `Table 1 median` = c(83, 72),
  `Table 1 IQR` = c("73-95", "61-84")
) |>
  knitr::kable(caption = "Virtual cohort covariates against Beijer 2026 Table 1.")

# The draw must reproduce the Table 1 centre. n = 200 gives a standard error of
# about 2% on a median, so 8% is loose enough for any thread count and still
# breaks on a mis-specified distribution.
stopifnot(
  abs(median(subjects$WT)   - 83) / 83 < 0.08,
  abs(median(subjects$CRCL) - 72) / 72 < 0.08
)
```

The three regimens compared are the two currently recommended ones (Figure 3b
and 3c: 2 g over 20 min or over 30 min, repeated at the guideline 2 h interval)
and the proposed alternative (Figure 3d: a 1 g loading dose over 20 min followed
by a continuous 1 g/h infusion).

Each regimen is simulated on the **same** virtual patients, with the same
between-subject random effects, so the comparison between regimens is paired
the way the paper's is (it re-simulated its own 200 patients, and then its 1000
virtual patients, under each regimen). That is achieved by giving every arm the
same subject IDs and resetting rxode2's simulation seed before each solve, which
gives common random numbers across arms. The assertion after the solve checks
that it actually worked rather than assuming it: individual clearance must come
out bit-identical in all three arms.

```{r events}
obs_times <- seq(0, 6, by = 1 / 120)

make_arm <- function(label) {
  per_subject <- lapply(seq_len(nrow(subjects)), function(i) {
    s  <- subjects[i, ]
    id <- s$subj
    if (label == "1 g/20 min + 1 g/h CI") {
      # Loading dose over 20 min, then a 6 h continuous infusion at 1 g/h
      # expressed as amt = 6000 mg at rate = 1000 mg/h.
      dose <- data.frame(
        time = c(0, 20 / 60), amt = c(1000, 6000),
        dur = c(20 / 60, NA_real_), rate = c(NA_real_, 1000)
      )
    } else {
      dur <- if (label == "2 g/20 min q2h") 20 / 60 else 30 / 60
      dose <- data.frame(
        time = c(0, 2), amt = 2000, dur = dur, rate = NA_real_
      )
    }
    dose$cmt <- "central"; dose$evid <- 1L; dose$dvid <- NA_integer_
    obs <- data.frame(
      time = obs_times, amt = NA_real_, dur = NA_real_, rate = NA_real_,
      cmt = "central", evid = 0L, dvid = 1L
    )
    out <- rbind(dose, obs)
    out <- out[order(out$time, -out$evid), ]
    out$id <- id; out$WT <- s$WT; out$CRCL <- s$CRCL; out$regimen <- label
    out
  })
  dplyr::bind_rows(per_subject)
}

regimens <- c("2 g/20 min q2h", "2 g/30 min q2h", "1 g/20 min + 1 g/h CI")
arms <- lapply(regimens, make_arm)
names(arms) <- regimens

# Within each arm the IDs must be unique; they are deliberately REUSED across
# arms so the regimens share subjects and random effects. Each arm is therefore
# solved in its own rxSolve call -- putting all three in one event table would
# merge each ID's three regimens into one subject receiving the summed dose.
for (a in arms) stopifnot(!anyDuplicated(unique(a[, c("id", "time", "evid")])))
stopifnot(vapply(arms, \(a) length(unique(a$id)), integer(1)) == n_per_arm)
```

```{r simulate}
sim <- dplyr::bind_rows(lapply(regimens, function(r) {
  # Reset rxode2's simulation seed before each arm so all three arms draw the
  # SAME between-subject random effects (common random numbers). This fixes the
  # draw within a render; it does not and cannot fix it across rxode2 versions
  # or solver thread counts, which is why the gates below are written to hold
  # for any cohort the model can produce.
  rxode2::rxSetSeed(20260913)
  rxode2::rxSolve(mod, events = arms[[r]], keep = c("regimen", "WT", "CRCL")) |>
    as.data.frame()
}))

# A large log-additive residual error can drive simulated concentrations very
# low, but the structural prediction Cc must stay non-negative.
stopifnot(all(sim$Cc >= 0), !anyNA(sim$Cc), !anyNA(sim$Ctot))

# Common random numbers actually held: individual clearance is a deterministic
# function of the covariates and etalcl, so bit-identical cl across arms proves
# every arm drew the same etas for the same subject. Without this the 20-min and
# 30-min arms would differ by two independent sampling errors (about 4
# percentage points on the target-attainment proportions) on top of the roughly
# 2-point real difference, and could even order the wrong way round.
cl_by_arm <- sim |>
  group_by(regimen, id) |>
  summarise(cl = dplyr::first(cl), .groups = "drop") |>
  pivot_wider(names_from = regimen, values_from = cl)
stopifnot(nrow(cl_by_arm) == n_per_arm)
cl_mat <- as.matrix(cl_by_arm[, regimens])
stopifnot(max(apply(cl_mat, 1, \(x) diff(range(x)) / mean(x))) < 1e-10)
```

# Figure 3 / Figure 4 -- target attainment

```{r figure-3, echo = FALSE}
sim |>
  group_by(regimen, time) |>
  summarise(
    Q05 = stats::quantile(Cc, 0.05), Q50 = stats::median(Cc),
    Q95 = stats::quantile(Cc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "#c0392b") +
  geom_line(colour = "#c0392b", linewidth = 0.8) +
  geom_hline(yintercept = c(1, 2), linetype = "dashed", colour = "grey30") +
  geom_vline(xintercept = 2, linetype = "dotted") +
  facet_wrap(~regimen) +
  scale_y_log10() +
  labs(
    x = "Time after start of the preoperative dose (h)",
    y = "Unbound cloxacillin (mg/L)",
    title = "Figure 3b-3d - unbound profiles by regimen (median, 5th-95th percentile)",
    caption = paste(
      "Replicates panels b, c and d of Figure 3 of Beijer 2026.",
      "Dashed lines are the 2 and 1 mg/L targets; dotted line is the guideline 2 h interval."
    )
  )
```

The paper's target-attainment criterion is an unbound plasma concentration
sustained at or above the MIC target "for at least 2 h after starting the
preoperative dose", and the proportions it reports are read off the
concentration 2 h after the start of the infusion.

The interval minimum has to be taken over the window **after** the loading
infusion completes, not from time zero: every arm starts at a structural zero
and climbs during its infusion, so a minimum taken from t = 0 would just return
the first grid point for every regimen and every arm would "fail". The longest
loading infusion here is 30 min, so the window below starts at 0.5 h. With that
window the two intermittent arms decline monotonically to the 2 h trough, which
makes the interval minimum equal to the 2 h concentration -- the check below
confirms that identity rather than assuming it, and it is what ties this
measure back to the paper's reported numbers. The continuous-infusion arm is
still rising at 2 h, so for that arm the two differ and the minimum is the
conservative measure.

```{r target-attainment}
# Window starts at 0.5 h, after the longest loading infusion (30 min) has
# finished. See the prose above for why a window starting at t = 0 is wrong.
ta <- sim |>
  filter(time >= 0.5, time <= 2) |>
  group_by(regimen, id) |>
  summarise(
    cmin_0_2 = min(Cc),
    c_at_2h  = Cc[which.min(abs(time - 2))],
    .groups  = "drop"
  )

# For the intermittent arms the post-infusion decline is monotone, so the
# interval minimum IS the 2 h value; for the continuous-infusion arm it is not
# (the profile is still rising towards its 13.1 mg/L steady state).
intermittent <- ta |> filter(regimen != "1 g/20 min + 1 g/h CI")
stopifnot(max(abs(intermittent$cmin_0_2 - intermittent$c_at_2h)) < 1e-6)
ci_arm <- ta |> filter(regimen == "1 g/20 min + 1 g/h CI")
stopifnot(all(ci_arm$cmin_0_2 <= ci_arm$c_at_2h))

ta_summary <- ta |>
  group_by(regimen) |>
  summarise(
    `Below 2 mg/L (%)` = 100 * mean(cmin_0_2 < 2),
    `Below 1 mg/L (%)` = 100 * mean(cmin_0_2 < 1),
    `Median Cmin over 0.5-2 h (mg/L)` = stats::median(cmin_0_2),
    .groups = "drop"
  )

published_ta <- tibble::tribble(
  ~regimen,                ~`Published below 2 mg/L (%)`, ~`Published below 1 mg/L (%)`,
  "2 g/20 min q2h",        26,                            6,
  "2 g/30 min q2h",        24,                            5,
  "1 g/20 min + 1 g/h CI", 0,                             0
)

ta_summary |>
  left_join(published_ta, by = "regimen") |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  rename(Regimen = regimen) |>
  knitr::kable(
    caption = paste(
      "Model-predicted target attainment against the 1000-simulated-patient",
      "results in the Beijer 2026 Results section ('Model-predicted target",
      "attainment in simulated patients') and Figure 4."
    )
  )
```

```{r target-attainment-gates}
pct <- function(reg, target) {
  v <- ta_summary[[paste0("Below ", target, " mg/L (%)")]][ta_summary$regimen == reg]
  if (length(v) != 1L) stop("no unique row for regimen '", reg, "'")
  v
}

# n = 200 gives a standard error of about 3 percentage points on a 25%
# proportion, so a 10-point band admits the cohort noise at any thread count
# while still going red on a mis-transcribed clearance, volume or dose -- those
# move these proportions by tens of points.
stopifnot(
  abs(pct("2 g/20 min q2h", 2) - 26) < 10,
  abs(pct("2 g/30 min q2h", 2) - 24) < 10,
  abs(pct("2 g/20 min q2h", 1) -  6) < 8,
  abs(pct("2 g/30 min q2h", 1) -  5) < 8
)

# Because the arms share subjects and random effects, the 20-vs-30-min
# comparison is PAIRED and deterministic: spreading the same 2 g over 30 min
# instead of 20 min delays the mass entering the central compartment, so every
# subject's 2 h concentration is higher and none is lower. This is a structural
# property of the ODE system for a fixed eta draw, not a cohort statistic, so
# the strict per-subject inequality is the right assertion -- and it is what
# recovers the direction of the paper's own 26% vs 24%.
paired <- ta |>
  filter(regimen != "1 g/20 min + 1 g/h CI") |>
  select(regimen, id, c_at_2h) |>
  pivot_wider(names_from = regimen, values_from = c_at_2h)
stopifnot(nrow(paired) == n_per_arm)
stopifnot(all(paired[["2 g/30 min q2h"]] > paired[["2 g/20 min q2h"]]))
stopifnot(pct("2 g/30 min q2h", 2) <= pct("2 g/20 min q2h", 2))
cat(sprintf(
  "Paired 30-min-minus-20-min gain in the 2 h unbound concentration: median %.3f mg/L (all %d subjects positive)\n",
  stats::median(paired[["2 g/30 min q2h"]] - paired[["2 g/20 min q2h"]]), n_per_arm
))

# The paper's central claim about the proposed regimen: a 1 g/h continuous
# infusion after a 1 g loading dose holds MORE than 99% of patients above both
# targets for the duration of the infusion. Asserted at 97% rather than 100% so
# the gate admits cohort noise but still fails if the loading dose or the
# infusion rate is wrong -- either breaks attainment for a large fraction.
ci_above_both <- 100 - pct("1 g/20 min + 1 g/h CI", 2)
cat(sprintf("Continuous-infusion arm above 2 mg/L throughout 0.5-2 h: %.1f%%   [paper: >99%%]\n",
            ci_above_both))
stopifnot(ci_above_both >= 97)

# And the paper's median 2 h concentration under that regimen.
ci_med_2h <- stats::median(ta$c_at_2h[ta$regimen == "1 g/20 min + 1 g/h CI"])
cat(sprintf("Continuous-infusion arm median unbound at 2 h: %.1f mg/L   [paper: 10.5 (3.3-38.6)]\n",
            ci_med_2h))
# Cohort median, so a 35% band; the paper's own range spans an order of
# magnitude and the typical-value gate above pins this to 10.6 exactly.
stopifnot(abs(ci_med_2h - 10.5) / 10.5 < 0.35)
```

The reproduction is close on both intermittent arms and on the continuous
infusion. Note that the paper reports two different sets of proportions for the
intermittent regimens: 22% and 18% below 2 mg/L for the **200 study patients**
re-simulated on their own observed covariates and empirical Bayes estimates
(Figure 3b, 3c), against 26% and 24% for the **1000 virtual patients** drawn
from the population model (Figure 4). The virtual-patient figures are the right
comparison for this vignette, because the cohort here is drawn from the
population distributions rather than from the study's individual estimates; the
virtual-patient proportions are higher precisely because they carry the full
between-subject variance instead of shrunken individual estimates.

Because the three arms share subjects and random effects, the 20-min-versus-
30-min contrast is a paired one and comes out in the paper's direction: the
longer infusion delays the mass entering the central compartment, so the 2 h
concentration is higher for every single subject. The absolute size of that
gain is small, which is exactly the paper's point -- lengthening the infusion
from 20 to 30 min buys only a couple of percentage points of target attainment,
whereas switching to a continuous infusion buys essentially all of it.

# PKNCA validation

The paper reports no NCA table, so there is no published Cmax / AUC /
half-life to compare against and `ncaComparisonTable()` has no reference to
consume. What is available instead is a stronger, exact check: because the
disposition is linear in unbound drug, unbound clearance times unbound
AUC to infinity must equal the administered dose for every subject
individually. That identity is what confirms the dose enters `central`
unscaled, which is the one structural reading of Figure S1 that the paper never
writes out in words.

```{r pknca-events}
# A single 2 g dose over 20 min, on the same virtual cohort, with a grid fine
# enough near Tmax that trapezoidal AUC is not biased low, and long enough that
# the extrapolated tail is negligible (the terminal half-life is about 1.7 h,
# so 24 h is roughly 14 half-lives).
nca_times <- sort(unique(c(
  seq(0, 1, by = 1 / 120),
  seq(1, 6, by = 0.05),
  seq(6, 24, by = 0.25)
)))

nca_events <- dplyr::bind_rows(lapply(seq_len(nrow(subjects)), function(i) {
  s <- subjects[i, ]
  dose <- data.frame(time = 0, amt = 2000, dur = 20 / 60, cmt = "central",
                     evid = 1L, dvid = NA_integer_)
  obs  <- data.frame(time = nca_times, amt = NA_real_, dur = NA_real_,
                     cmt = "central", evid = 0L, dvid = 1L)
  out <- rbind(dose, obs)
  out <- out[order(out$time, -out$evid), ]
  out$id <- s$subj; out$WT <- s$WT; out$CRCL <- s$CRCL
  out
}))

# zeroRe() here so the NCA operates on each subject's structural profile: the
# CL x AUC identity is a property of the ODE system, and log-additive residual
# error would add noise to AUC without testing anything about the model.
sim_nca_raw <- rxode2::rxSolve(
  rxode2::zeroRe(mod), events = nca_events,
  keep = c("WT", "CRCL")
) |>
  as.data.frame()

# Solver noise in a 14-half-life tail can dip slightly negative; PKNCA then
# takes log() of a negative value and aucinf.obs comes back NaN.
stopifnot(all(sim_nca_raw$Cc >= 0))
```

```{r pknca}
# Only `!is.na(Cc)`: a `time > 0` or `Cc > 0` filter would drop the time-zero
# row that PKNCA needs to anchor AUC from 0, warning once per subject.
sim_nca <- sim_nca_raw |>
  filter(!is.na(Cc)) |>
  mutate(regimen = "2 g/20 min single dose") |>
  select(id, time, Cc, regimen)

# Guarantee a time = 0 record per subject; pre-dose unbound cloxacillin is 0.
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, regimen) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, regimen, time, .keep_all = TRUE) |>
  arrange(id, regimen, time)

conc_obj <- PKNCA::PKNCAconc(
  sim_nca, Cc ~ time | regimen + id,
  concu = "mg/L", timeu = "h"
)

dose_df <- nca_events |>
  filter(evid == 1L) |>
  mutate(regimen = "2 g/20 min single dose") |>
  select(id, time, amt, dur, regimen)

dose_obj <- PKNCA::PKNCAdose(
  dose_df, amt ~ time | regimen + id,
  duration = "dur", route = "intravascular", doseu = "mg"
)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, auclast = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)

nca_wide <- as.data.frame(nca_res) |>
  select(id, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
stopifnot(nrow(nca_wide) == n_per_arm, !anyNA(nca_wide$aucinf.obs))
```

```{r pknca-mass-balance}
# Per-subject unbound clearance, as returned by rxSolve.
cl_per_id <- sim_nca_raw |> group_by(id) |> summarise(cl = dplyr::first(cl), .groups = "drop")

mb <- nca_wide |>
  left_join(cl_per_id, by = "id") |>
  mutate(recovered = cl * aucinf.obs, pct_diff = 100 * (recovered - 2000) / 2000)

tibble::tibble(
  Check = "CL * AUCinf / dose - 1 (%)",
  Median = round(stats::median(mb$pct_diff), 3),
  `90th pct of |diff|` = round(stats::quantile(abs(mb$pct_diff), 0.9), 3),
  Max = round(max(abs(mb$pct_diff)), 3)
) |>
  knitr::kable(caption = "Mass balance: unbound clearance times unbound AUC to infinity recovers the 2 g dose.")

# Both sides of this comparison use the SAME drawn parameters -- it is a solve
# against its own closed form, so the residual is pure trapezoidal and
# extrapolation error, not per-subject physiology. A tight bound is correct here
# and should be kept; it is what proves the dose enters `central` unscaled with
# no bioavailability term.
stopifnot(max(abs(mb$pct_diff)) < 0.5)
```

```{r pknca-summary}
nca_summary <- nca_wide |>
  summarise(across(
    c(cmax, tmax, auclast, aucinf.obs, half.life),
    list(median = \(x) stats::median(x),
         p05 = \(x) stats::quantile(x, 0.05),
         p95 = \(x) stats::quantile(x, 0.95))
  )) |>
  pivot_longer(everything(), names_to = c("param", "stat"), names_sep = "_") |>
  pivot_wider(names_from = stat, values_from = value) |>
  mutate(
    Parameter = c(
      cmax = "Cmax (mg/L)", tmax = "Tmax (h)", auclast = "AUClast (mg*h/L)",
      aucinf.obs = "AUC0-inf (obs) (mg*h/L)", half.life = "t1/2 (h)"
    )[param],
    across(c(median, p05, p95), \(x) signif(x, 3))
  ) |>
  select(Parameter, Median = median, `5th pct` = p05, `95th pct` = p95)

knitr::kable(
  nca_summary,
  caption = paste(
    "Unbound-cloxacillin NCA over the virtual cohort after a single 2 g dose",
    "over 20 min. Beijer 2026 reports no NCA table, so these values are",
    "descriptive; the mass-balance check above is the validating gate."
  )
)

# Free regression tests from the model's own structure. Tmax must land at the
# end of the 20 min infusion, and the terminal half-life must match the smaller
# eigenvalue of the typical-value two-compartment system.
kel_t <- 76.4 / 57.1; k12_t <- 57.2 / 57.1; k21_t <- 57.2 / 69.0
a <- kel_t + k12_t + k21_t
lambda_z <- (a - sqrt(a^2 - 4 * kel_t * k21_t)) / 2
t_half_typical <- log(2) / lambda_z
cat(sprintf("Typical-value terminal half-life from the eigenvalue: %.3f h\n", t_half_typical))
cat(sprintf("Cohort median PKNCA half-life: %.3f h\n", stats::median(nca_wide$half.life)))

stopifnot(
  # Deterministic: every subject peaks at the end of the infusion.
  max(abs(nca_wide$tmax - 20 / 60)) < 1e-6,
  # AUClast must be just under AUCinf after 14 half-lives, never over it.
  all(nca_wide$auclast <= nca_wide$aucinf.obs),
  all(nca_wide$auclast / nca_wide$aucinf.obs > 0.99),
  # The cohort median half-life sits near the typical-value eigenvalue. IIV on
  # CL, V1 and Q spreads the individual eigenvalues, so this is a 25% band on a
  # cohort median rather than a tight match.
  abs(stats::median(nca_wide$half.life) - t_half_typical) / t_half_typical < 0.25
)
```

# Assumptions and deviations

- **The ODE states hold unbound drug, and the binding term is an output
  transformation.** Beijer 2026 never writes the differential equations out;
  Figure S1 draws them. The reading encoded here -- `CL` and `Q` acting on the
  unbound pool, `V1` and `V2` as unbound volumes, the bound plasma pool as an
  algebraic equilibrium off `V1` that moves no mass, and the full 2000 mg
  entering `central` with no bioavailability term -- is what reproduces the
  paper's own reported numbers: the typical-patient 2 h unbound concentration
  (3.23 vs 3.2 mg/L), the continuous-infusion 2 h median (10.6 vs 10.5 mg/L),
  the observed start-of-surgery median (10.9 vs 10 mg/L) and the observed median
  protein binding (90.5 vs 91%). The mass-balance gate confirms the dose scaling
  exactly. A mass-conserving alternative, in which the dose is distributed
  between a bound and an unbound plasma pool, would put the unbound
  concentrations an order of magnitude lower and reproduces none of those.
- **Covariate centring constants come from Table S2's printed equation**, not
  from the Table 1 cohort medians. Table S2's CL row reads `(BW/84)` and
  `(eGFR/67)` while Table 1 gives medians of 83 kg and 72 mL/min/1.73 m^2. The
  Figure 1 caption independently names "a body weight of 84 kg and eGFR
  67 mL/min/1.73 m2" as the typical patient, which corroborates the equation.
  This is the same printed-equation-beats-demographics-table pattern already
  recorded for several other models in this library.
- **Omega values are log-scale standard deviations, squared into `ini()`
  variances.** Settled by Table S2's own CV column rather than assumed -- see
  the "Omega scale" chunk above.
- **`CRCL` carries the Lund-Malmo Revised 2018 relative (BSA-normalized) eGFR.**
  Table S1 shows the paper evaluated CKD-EPI 2021, Cockcroft-Gault
  (original and weight-adjusted) and LMR18 in relative and absolute forms;
  LMR18 correlated best. A user supplying a CKD-EPI eGFR, a raw (non-BSA-
  normalized) clearance, or the absolute LMR18 estimate is not supplying the
  column this model was fitted on. The absolute LMR18 estimate correlated even
  better with `eta_CL` (r = 0.43 against 0.32) but was rejected because it left
  every anthropometric covariate independently correlated with clearance.
- **Plasma albumin is documented but not used.** It was significantly correlated
  with protein binding, but the Supplementary Material states that adding it as
  a covariate on `Bmax` (and `Kd`) "led to over-fitting and did not improve
  overall model performance". It is recorded in `covariatesDataExcluded`
  alongside age, sex, height, BMI, BSA, creatinine, ASA class and arthroplasty
  site, all screened and not retained.
- **Virtual-cohort covariates are drawn independently.** Body weight and
  relative eGFR are log-normal fits to the Table 1 median and IQR, truncated to
  the Table 1 range, and drawn without correlation. The paper drew from
  "distributions assumed to be identical to the covariate distributions in our
  study population" without stating a correlation structure either. Since
  relative eGFR is BSA-normalized, its residual correlation with body weight is
  weaker than it would be for an absolute estimate.
- **The cohort is 200 per arm, not the paper's 1000.** The library caps
  simulation cohorts at 200 per arm. This widens the sampling noise on the
  target-attainment proportions to roughly 3 percentage points, which is why
  those gates carry a 10-point band while the deterministic typical-value gates
  are tight.
- **The three regimens share subjects and random effects (common random
  numbers).** Every arm uses the same subject IDs and `rxode2::rxSetSeed()` is
  reset before each arm's solve, which is what makes the regimen contrasts
  paired, as the paper's were. The `simulate` chunk asserts that this held by
  checking individual clearance is bit-identical across arms. Without it the two
  intermittent arms carry two independent sampling errors on top of a roughly
  2-percentage-point real difference, and can order the wrong way round.
- **Target attainment is compared against the virtual-patient figures (26% and
  24%), not the study-patient figures (22% and 18%).** The paper reports both;
  the virtual-patient set is the like-for-like comparison, as explained in the
  target-attainment section. The two intermittent arms differ by only 2
  percentage points in the paper, which a 200-subject cohort cannot resolve, so
  no gate attempts to.
- **No published NCA comparison table.** Beijer 2026 reports no Cmax, Tmax, AUC
  or half-life, so `ncaComparisonTable()` is not used. The PKNCA output is
  descriptive and the validating gates are the exact `CL * AUCinf = dose` mass
  balance, the `Tmax` identity, and the terminal-half-life eigenvalue check.
- **The model carries little information about severe renal impairment or very
  low albumin.** The paper says so explicitly (Strengths and limitations):
  relative eGFR ranged 18-142 mL/min/1.73 m^2 but with few patients at very low
  levels, and albumin was generally well preserved.
- **No epidemiological cut-off value exists for cloxacillin against
  *S. aureus*.** The 2 mg/L primary and 1 mg/L secondary targets are borrowed
  from oxacillin's established ECOFF and flucloxacillin's tentative one; the
  paper flags this and notes that the single available EUCAST MIC distribution
  suggests 0.5-1 mg/L would be plausible for cloxacillin. The targets drawn on
  the figures above are the paper's, not a property of the model.
