---
title: "Cysteamine (Belldina 2003)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cysteamine (Belldina 2003)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
suppressPackageStartupMessages({
  library(nlmixr2lib)
  library(PKNCA)
  library(rxode2)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Belldina_2003_cysteamine"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Belldina_2003_cysteamine"))$description`
* Article: <https://doi.org/10.1046/j.1365-2125.2003.01927.x>

# Population

The model was developed at the University of California, San Diego from 11
paediatric and young-adult patients (4 female, 7 male; 10 Caucasian, 1
Caucasian/Hispanic) aged 3-15 years with body weight 14.3-60.2 kg and height
93.0-162.0 cm, all with nephropathic cystinosis without renal transplant and
on chronic cysteamine bitartrate (Cystagon) therapy for at least 12 months
prior to the study (Belldina 2003 Table 1). The single-dose, open-label,
steady-state study administered each patient's regular cysteamine bitartrate
dose (225-550 mg, approximately 0.05 mmol/kg, given every six hours) at
approximately 08.00 h with 100 mL of ambient-temperature water, with a
standardised low-fat / low-protein breakfast 30 min before dosing and a
standardised lunch 4 h post-dose. Blood samples for plasma cysteamine were
collected pre-dose and at 0.5, 1, 1.5, 2, 3, and 6 h post-dose; the same
samples provided white-blood-cell preparations for cystine content assay.

The same information is available programmatically via
`readModelDb("Belldina_2003_cysteamine")$population`.

# Source trace

Per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Belldina_2003_cysteamine.R`. The table
below collects them for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Structural PK | 2-compartment oral, first-order absorption, absorption lag | Belldina 2003 Methods "Data analysis", page 521-522 |
| Structural PD | Fractional inhibitory Emax (Hill = 1) on effect-compartment Ce | Belldina 2003 Methods "Data analysis", Eq. 1 |
| `lka` (Ka) | `log(1.7)` 1/h | Table 2: Ka mean = 1.7 1/h (range 0.7-2.5; geomean 1.6) |
| `lcl` (CL/F) | `log(70.74)` L/h | Table 2: CL/F mean = 32.3 mL/min/kg * 36.5 kg * 60/1000 |
| `lvc` (Vc/F) | `log(73.0)` L | Table 2: Vc mean = 2.0 L/kg * 36.5 kg |
| `lq` (Q/F) | `log(65.26)` L/h | Table 2: Q mean = 29.8 mL/min/kg * 36.5 kg * 60/1000 |
| `lvp` (Vp/F) | `log(478.15)` L | Table 2: Vss/F mean = 15.1 L/kg * 36.5 kg = 551.15 L; Vp = Vss - Vc |
| `ltlag` (Alag) | `log(0.44)` h | Table 2: Alag mean = 0.44 h (range 0.22-0.92; geomean 0.41) |
| `e_wt_cl_q`, `e_wt_vc_vp` | `fixed(1.0)` | Paper per-kg parameterisation -> linear weight scaling |
| `lec50` (EC50) | `log(15.3)` umol/L | Table 3: EC50 mean = 15.3 uM (range 0.6-61.1; geomean 5.6) |
| `lke0` (Ke0) | `log(2.2)` 1/h | Table 3: Ke0 mean = 2.2 1/h (range 0.2-8.9; geomean 1.3) |
| `lbl` (BL) | `log(0.91)` nmol/mg | Table 3: BL mean = 0.91 nmol/mg (range 0.13-1.9; geomean 0.76) |
| `etalka` IIV variance | 0.121 | 2*log(1.7/1.6); Table 2 mean/geomean |
| `etalcl` IIV variance | 0.108 | 2*log(32.3/30.6); Table 2 mean/geomean |
| `etalvc` IIV variance | 0.446 | 2*log(2.0/1.6); Table 2 mean/geomean |
| `etalq` IIV variance | 0.384 | 2*log(29.8/24.6); Table 2 mean/geomean |
| `etalvp` IIV variance | 0.634 | 2*log(15.1/11.0); Table 2 mean/geomean (Vss proxy for Vp) |
| `etaltlag` IIV variance | 0.141 | 2*log(0.44/0.41); Table 2 mean/geomean |
| `etalec50` IIV variance | 2.011 | 2*log(15.3/5.6); Table 3 mean/geomean (very large spread) |
| `etalke0` IIV variance | 1.052 | 2*log(2.2/1.3); Table 3 mean/geomean |
| `etalbl` IIV variance | 0.360 | 2*log(0.91/0.76); Table 3 mean/geomean |
| `propSd` (plasma) | 0.15 (assumed) | not reported in Belldina 2003 |
| `addSd_cystine` (PD) | 0.10 (assumed) | not reported in Belldina 2003 |
| `d/dt(depot) = -ka * depot` | n/a | First-order absorption from gut |
| `d/dt(central) = ka*depot - kel*central - k12*central + k21*peripheral1` | n/a | Two-compartment disposition |
| `d/dt(peripheral1) = k12*central - k21*peripheral1` | n/a | Peripheral mass-balance |
| `alag(depot) = tlag` | n/a | Absorption lag |
| `Cc = central / vc * 1000 / 227.24` | n/a | Salt-mass -> uM cysteamine free base (1:1 stoichiometry) |
| `d/dt(effect) = ke0 * (Cc - effect)` | n/a | Effect-compartment link (Sheiner-Holford-Stanski-Pieper) |
| `cystine = bl * ec50 / (ec50 + effect)` | n/a | Eq. 1: fractional inhibitory Emax (Hill = 1) |

# Virtual cohort

Individual-level data from the 11-patient cohort are not publicly available.
The virtual cohort below approximates the demographic spread reported in
Belldina 2003 Table 1 by drawing body weight uniformly across the reported
range and assigning each subject a cysteamine bitartrate dose of
approximately 11 mg/kg (the paper's typical 0.05 mmol/kg every-6-hour
schedule, with MW_salt = 227.24 g/mol). Each subject receives five doses to
reach the steady-state condition described in the paper; the dosing interval
under study is the last (steady-state) 6 h window.

```{r cohort}
set.seed(20030428)
n_subj <- 200L

cohort <- tibble(
  id        = seq_len(n_subj),
  WT        = runif(n_subj, min = 14.3, max = 60.2),    # Belldina 2003 Table 1 range
  treatment = factor("Cysteamine bitartrate 11.4 mg/kg PO Q6H")
) |>
  dplyr::mutate(
    dose_mg = round(11.4 * WT, 1)                       # ~ 0.05 mmol/kg salt
  )

stopifnot(!anyDuplicated(cohort$id))
```

# Simulation

The dosing schedule is `ii = 6, addl = 4` (5 doses total across 24 h) into
the `depot` compartment, with steady state taken at the last 6 h interval
(18-24 h). Observations are taken on a dense grid spanning the steady-state
interval so the simulated profile is directly comparable with Belldina 2003
Figures 1A (cysteamine plasma) and 1B (WBC cystine).

```{r events}
ss_start <- 18                                          # last dose at hour 18
ss_end   <- 24
obs_grid <- sort(unique(c(seq(0, ss_end, by = 0.25),
                          seq(ss_start, ss_end, by = 0.1))))

dose_rows <- cohort |>
  dplyr::mutate(
    time = 0,
    amt  = dose_mg,
    cmt  = "depot",
    evid = 1L,
    ii   = 6,
    addl = 4L
  )

# Schedule only Cc observations -- both Cc (plasma cysteamine) and cystine
# (WBC cystine PD output) are produced at every observation time as columns
# of the rxSolve return value, so there is no need to schedule separate
# event-table rows for each output.
obs_rows <- cohort |>
  tidyr::crossing(time = obs_grid) |>
  dplyr::mutate(amt = NA_real_, cmt = "Cc", evid = 0L,
                ii = NA_real_, addl = NA_integer_)

events <- dplyr::bind_rows(dose_rows, obs_rows) |>
  dplyr::select(id, time, amt, cmt, evid, ii, addl, WT, dose_mg, treatment) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid", "cmt")])))
```

```{r simulate}
mod <- rxode2::rxode2(readModelDb("Belldina_2003_cysteamine"))
sim <- rxode2::rxSolve(
  mod, events = events,
  keep = c("WT", "dose_mg", "treatment"),
  returnType = "data.frame"
) |>
  dplyr::filter(!is.na(Cc))   # drop dosing rows; observation rows have populated Cc
```

# Replicate published figures

## Figure 1A: typical-value plasma cysteamine over a steady-state interval

Belldina 2003 Figure 1A plots the mean +/- SD plasma cysteamine concentration
across the 6 h dosing interval at steady state, in 11 patients. Reported
mean Cmax = 36.3 +/- 11.7 uM (range 16.9-53.2 uM) at Tmax ~ 1.4 h (Results).
The block below zero-out random effects and plots a typical 36.5 kg patient
receiving 400 mg of cysteamine bitartrate every 6 hours, with the last
(steady-state) interval rendered relative to its administration time.

```{r figure-1A}
mod_typical <- mod |> rxode2::zeroRe()

typical_cohort <- tibble(
  id        = 1L,
  WT        = 36.5,                                  # arithmetic mean of N = 11
  dose_mg   = 400,                                   # ~ median of Table 1 doses
  treatment = factor("Typical 36.5 kg patient, 400 mg PO Q6H")
)
typical_dose <- typical_cohort |>
  dplyr::mutate(time = 0, amt = dose_mg, cmt = "depot", evid = 1L,
                ii = 6, addl = 4L)
typical_obs <- typical_cohort |>
  tidyr::crossing(time = obs_grid) |>
  dplyr::mutate(amt = NA_real_, cmt = "Cc", evid = 0L,
                ii = NA_real_, addl = NA_integer_)
typical_events <- dplyr::bind_rows(typical_dose, typical_obs) |>
  dplyr::select(id, time, amt, cmt, evid, ii, addl, WT, dose_mg, treatment) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim_typical <- rxode2::rxSolve(
  mod_typical, events = typical_events,
  keep = c("WT", "dose_mg", "treatment"),
  returnType = "data.frame"
) |>
  dplyr::filter(!is.na(Cc))

typical_cc_ss <- sim_typical |>
  dplyr::filter(time >= ss_start, time <= ss_end) |>
  dplyr::transmute(rel_time = time - ss_start, Cc)

ggplot(typical_cc_ss, aes(rel_time, Cc)) +
  geom_line(linewidth = 1.0) +
  labs(x = "Time within steady-state dosing interval (h)",
       y = "Plasma cysteamine (umol/L)",
       title = "Replicates Figure 1A: plasma cysteamine at steady state",
       caption = "Typical 36.5 kg patient, 400 mg cysteamine bitartrate Q6H. Zero random effects.")
```

## Figure 1B: typical-value WBC cystine over a steady-state interval

Belldina 2003 Figure 1B plots the mean +/- SD WBC cystine content across
the 6 h interval. The Results section reports a maximum decrement of
0.46 +/- 0.23 nmol cystine / mg protein (range 0.07-0.81) at typical time
1.8 +/- 0.8 h, i.e. roughly 50% reduction from baseline 0.91 nmol/mg.

```{r figure-1B}
typical_cystine_ss <- sim_typical |>
  dplyr::filter(time >= ss_start, time <= ss_end) |>
  dplyr::transmute(rel_time = time - ss_start, cystine)

ggplot(typical_cystine_ss, aes(rel_time, cystine)) +
  geom_line(linewidth = 1.0) +
  geom_hline(yintercept = 0.91, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 5, y = 0.93, label = "Baseline BL = 0.91",
           colour = "grey40", size = 3.2, hjust = 1) +
  labs(x = "Time within steady-state dosing interval (h)",
       y = "WBC cystine content (nmol cystine / mg protein)",
       title = "Replicates Figure 1B: WBC cystine at steady state",
       caption = "Typical 36.5 kg patient, 400 mg cysteamine bitartrate Q6H. Zero random effects.")
```

```{r figure-1-summary}
peak_plasma <- typical_cc_ss |>
  dplyr::summarise(Cmax = max(Cc, na.rm = TRUE),
                   Tmax = rel_time[which.max(Cc)])
min_cystine <- typical_cystine_ss |>
  dplyr::summarise(min_cyst = min(cystine, na.rm = TRUE),
                   t_min    = rel_time[which.min(cystine)])
decrement <- 0.91 - min_cystine$min_cyst

knitr::kable(
  tibble::tibble(
    Quantity = c("Plasma Cmax (umol/L)",
                 "Plasma Tmax (h)",
                 "Minimum WBC cystine (nmol/mg)",
                 "Time to minimum (h)",
                 "Cystine decrement (nmol/mg)",
                 "Fractional cystine decrement"),
    `Belldina 2003` = c("36.3 +/- 11.7", "1.4", "(BL 0.91 - 0.46 = 0.45)",
                        "1.8 +/- 0.8", "0.46 +/- 0.23", "~ 47%"),
    Simulation = c(sprintf("%.1f", peak_plasma$Cmax),
                   sprintf("%.2f", peak_plasma$Tmax),
                   sprintf("%.2f", min_cystine$min_cyst),
                   sprintf("%.2f", min_cystine$t_min),
                   sprintf("%.2f", decrement),
                   sprintf("%.0f%%", 100 * decrement / 0.91))
  ),
  caption = "Typical-value peak / decrement metrics: simulation vs Belldina 2003 Results."
)
```

## Figure 2: counter-clockwise hysteresis (cystine vs plasma cysteamine)

Belldina 2003 Figure 2 plots WBC cystine content against plasma cysteamine
concentration in a temporal-loop fashion, showing the counter-clockwise
hysteresis pattern that is diagnostic of an effect-compartment delay
(Ke0). Reproduced here from the typical-value steady-state simulation; the
loop traces from baseline cystine downward as plasma cysteamine rises after
each dose, then upward back to baseline as plasma falls (lagged by ~ 1/Ke0
in time).

```{r figure-2}
hyst <- dplyr::inner_join(
  typical_cc_ss      |> dplyr::transmute(rel_time, Cc),
  typical_cystine_ss |> dplyr::transmute(rel_time, cystine),
  by = "rel_time"
)

ggplot(hyst, aes(Cc, cystine, colour = rel_time)) +
  geom_path(arrow = grid::arrow(length = grid::unit(0.15, "cm"), type = "closed"),
            linewidth = 0.9) +
  scale_colour_viridis_c(name = "Time (h)") +
  labs(x = "Plasma cysteamine concentration (umol/L)",
       y = "WBC cystine content (nmol cystine / mg protein)",
       title = "Replicates Figure 2: counter-clockwise hysteresis (cystine vs cysteamine)",
       caption = "Typical-value trajectory over a single steady-state dosing interval. Zero random effects.")
```

## Plasma profile across the virtual cohort

```{r stochastic-plasma}
sim |>
  dplyr::filter(time >= ss_start, time <= ss_end) |>
  dplyr::mutate(rel_time = time - ss_start) |>
  dplyr::group_by(rel_time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(rel_time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time within dosing interval (h)",
       y = "Plasma cysteamine (umol/L)",
       title = "Plasma cysteamine at steady state, virtual cohort",
       caption = "5-50-95 percentile envelope across 200 simulated subjects (WT 14.3-60.2 kg).")
```

# PKNCA validation

PKNCA is run on the simulated plasma profile across the last 6 h dosing
interval at steady state (steady-state recipe of `pknca-recipes.md`).
Reported NCA values are compared against Belldina 2003 Results.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(time >= ss_start, time <= ss_end) |>
  dplyr::transmute(id, time = time - ss_start, Cc, treatment)

# Steady-state dose at the start of the interval (taken as the actual mg
# administered per subject, not the per-kg derivation).
dose_df <- cohort |>
  dplyr::transmute(id, time = 0, amt = dose_mg, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "umol/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "mg")

intervals <- data.frame(
  start    = 0,
  end      = 6,
  cmax     = TRUE,
  tmax     = TRUE,
  auclast  = TRUE,
  cmin     = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- suppressWarnings(PKNCA::pk.nca(nca_data))
knitr::kable(summary(nca_res),
             caption = "Simulated steady-state NCA over 6 h dosing interval (n = 200).")
```

## Comparison against Belldina 2003 Results

```{r comparison}
nca_tbl <- as.data.frame(nca_res$result)

med_q <- function(test) {
  vals <- nca_tbl |>
    dplyr::filter(PPTESTCD == test) |>
    dplyr::pull(PPORRES) |>
    as.numeric()
  c(median = median(vals, na.rm = TRUE),
    q05    = quantile(vals, 0.05, na.rm = TRUE, names = FALSE),
    q95    = quantile(vals, 0.95, na.rm = TRUE, names = FALSE))
}

sim_cmax <- med_q("cmax")
sim_tmax <- med_q("tmax")

knitr::kable(
  tibble::tibble(
    Quantity = c("Cmax (umol/L)", "Tmax (h)"),
    `Belldina 2003 Results (mean +/- SD)` = c("36.3 +/- 11.7", "1.4 (range 1.0-2.0)"),
    `Simulated median (5-95%)` = c(
      sprintf("%.1f (%.1f-%.1f)", sim_cmax["median"], sim_cmax["q05"], sim_cmax["q95"]),
      sprintf("%.2f (%.2f-%.2f)", sim_tmax["median"], sim_tmax["q05"], sim_tmax["q95"])
    )
  ),
  caption = "Comparison: observed Belldina 2003 Results vs simulated 200-subject cohort."
)
```

# Assumptions and deviations

- **Two-stage analysis, not popPK.** Belldina 2003 estimated the
  pharmacokinetic and pharmacodynamic parameters as **individual NONMEM
  fits per patient** (Methods, page 521-522: "The pharmacokinetics of a
  two-compartment model with first-order absorption and a lag time were
  first determined in each individual using a proportional residual error
  model. These pharmacokinetic parameters were then fixed and the
  parameters of the pharmacodynamic model were estimated.") and summarised
  the 11 per-subject estimates as arithmetic mean / median / geometric
  mean / min / max in Tables 2 and 3. The paper does **not** report
  population OMEGA estimates. The library model encodes the **arithmetic
  means** as the typical values, in keeping with the precedent set by
  `Park_2001_ketoprofen.R` for the same individual-fits-summarised
  situation. Users wishing to use the geometric means as the typical
  values instead (Table 2 geometric means: CL/F 30.6 mL/min/kg, Vc 1.6
  L/kg, Q 24.6 mL/min/kg, Vss/F 11.0 L/kg, Ka 1.6 1/h, Alag 0.41 h;
  Table 3 geometric means: EC50 5.6 uM, Ke0 1.3 1/h, BL 0.76 nmol/mg)
  should override the `ini()` log-typical values accordingly.
- **IIV derived from mean / geometric-mean ratio.** Because no population
  OMEGAs were estimated, the IIV variances in the library model are
  derived from the cross-individual descriptive statistics using the
  log-normal identity `omega^2 = 2 * log(arithmetic mean / geometric
  mean)`. These variances **describe the spread of the 11 individual
  NONMEM fits**, not a formal popPK estimate of population variance.
  Users planning a refit should re-estimate these against new data.
- **Residual error magnitudes not reported.** PK Methods specify a
  proportional residual error structure but do not report its SD; PD
  Methods do not specify a residual error structure at all. The library
  assigns `propSd = 0.15` (15% proportional) on the plasma cysteamine
  output and `addSd_cystine = 0.10` nmol cystine/mg protein on the WBC
  cystine output, both clearly tagged as **assumed** in the `ini()`
  labels. Both are plausible magnitudes given the data range observed in
  the paper (Cmax 16.9-53.2 uM; cystine decrement 0.07-0.81 nmol/mg) but
  neither is paper-derived. A refit would re-estimate these.
- **Per-kg parameterisation -> linear weight scaling.** All clearance and
  volume terms in Table 2 are reported as per-kg quantities (CL/F in
  mL/min/kg, Vc/F and Vss/F in L/kg, Q in mL/min/kg), implying linear
  (allometric exponent = 1.0) weight scaling. The library encodes this
  via `e_wt_cl_q <- fixed(1.0)` and `e_wt_vc_vp <- fixed(1.0)` on the
  shared exponents, with absolute reference values computed at the
  arithmetic-mean weight 36.5 kg of the 11 patients in Table 1
  (43.8 + 39.0 + 57.2 + 22.5 + 18.4 + 14.3 + 29.1 + 47.0 + 60.2 + 38.7 +
  31.3 = 401.5 kg / 11). This differs from the canonical adult-popPK
  fixed-allometric scheme of 0.75 on clearances and 1.0 on volumes; it is
  the paper's explicit parameterisation.
- **Salt-form / molar conversion bridged inside `model()`.** Cysteamine
  bitartrate (Cystagon) is administered as the salt (MW 227.24 g/mol)
  but plasma is assayed as the cysteamine moiety (MW 77.15 g/mol, free
  base). Salt and moiety are present in 1:1 molar stoichiometry. The
  library accepts the dose record `amt` in **mg of bitartrate salt** (to
  match clinical practice and the paper's dose levels of 225-550 mg) and
  converts to micromolar cysteamine at the observation:
  `Cc <- central / vc * 1000 / 227.24`. This means `units$dosing = "mg"`
  is salt-mass while `units$concentration = "umol/L (uM)"` is free-base
  molar concentration; the convention checker flags the resulting
  mass-vs-molar dimensional mismatch on `units$dosing /
  units$concentration` as a warning, which is the intended behaviour for
  this model and should not be resolved by changing the units list.
- **Effect-compartment naming convention.** The paper labels the
  hypothetical effect-compartment concentration `Ce` and the elimination
  rate constant `Ke0`. The library uses the canonical compartment name
  `effect` (so `Ce` is the value of the `effect` state) and the canonical
  parameter name `lke0` / `ke0` (rate-constant form, 1/h, in line with
  `Park_2001_ketoprofen.R`'s `lkeo` precedent and the more recent
  `Kretsos_2014_olokizumab.R` / `Berges_2015_ozanezumab.R` /
  `Przybylowski_2015_propofol.R` precedents that standardise on `lke0`).
- **PD output name `cystine`.** Belldina 2003 reports the pharmacodynamic
  output as "WBC cystine content" with units nmol cystine / mg protein.
  The library names this output `cystine` (lower-case, paper-mechanistic
  output, exempt from the canonical `Cc_<metab>` naming rule per
  `compartment-names.md`) and applies an additive residual error
  `addSd_cystine`.
- **Bioavailability F absorbed into apparent CL/F and Vss/F.** Belldina
  2003 reports all clearance and volume quantities as the
  apparent-CL-or-V / F form because no IV reference was available; the
  library inherits this convention via `lcl` / `lvc` / `lq` / `lvp`
  representing absolute *apparent* (oral) values at the reference weight
  with no separately estimated bioavailability.
- **Hill exponent = 1 (no sigmoid Emax).** Belldina 2003 explicitly notes
  in Results that "Data were not sufficient to fit a sigmoid Emax model
  to the pharmacodynamic data." The library encodes Eq. 1 of the source
  paper directly as the fractional inhibitory Emax with Hill exponent
  implicit at 1: `cystine = BL * EC50 / (EC50 + Ce)`.
- **Virtual cohort body weight distribution.** The cohort of 200 subjects
  draws body weight uniformly across the paper's reported range
  (14.3-60.2 kg). Age, sex, and concomitant-medication burden are not
  model covariates and are not encoded.
- **Steady-state assumption.** The paper assumed steady state based on
  the patient self-reporting compliance and timing of the last two doses
  and the inter-dose interval. The library reproduces this by simulating
  five Q6H doses (24 h of accumulation) before evaluating the 18-24 h
  steady-state interval; the simulated steady-state Cmax for a 36.5 kg
  typical subject at the median 400 mg dose lands in the lower half of
  the observed range, reflecting the choice of a median weight + median
  dose rather than the heaviest-dose case.
