---
title: "Dalbavancin (Benavent 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dalbavancin (Benavent 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Benavent_2025_dalbavancin"))$reference`
* Article: [Antimicrob Agents Chemother 2025;69(12):e00773-25](https://doi.org/10.1128/aac.00773-25)
  (open access, CC BY 4.0; PMC12691635)

Dalbavancin is a lipoglycopeptide with a terminal half-life near two weeks, so a
single infusion can cover weeks of therapy. Benavent 2025 asked whether one
1500 mg dose is enough to finish the four-week antibiotic course that follows
first-stage surgery for a chronic prosthetic joint infection, where the infected
implant is removed, a vancomycin- plus gentamicin-loaded cement spacer is put in,
and a new prosthesis is implanted at a second operation months later. Twenty
patients were treated; eighteen contributed plasma concentrations to a population
PK model, which was then used for Monte Carlo target-attainment simulations.

The model is deliberately simple: one compartment, intravenous infusion,
first-order elimination, interindividual variability on both clearance and
volume, proportional residual error, and **no covariates** -- the covariate
search improved nothing, so the structural model is the final model.

```{r load-model}
# rxode2::rxode() resolves the model function to an rxUi without depending on
# ini()/model() being attached (see known-vignette-failure-patterns 6 and 7).
mod <- rxode2::rxode(readModelDb("Benavent_2025_dalbavancin"))
mod
```

# Population

Eighteen of the twenty enrolled patients contributed total plasma dalbavancin
concentrations, each supplying one to three samples drawn opportunistically at
weekly-to-biweekly outpatient visits through week four. The cohort is elderly
(median age 75.5 years, interquartile interval 69-79) and predominantly female
(70%, n = 14). Renal function and albumin were uniformly normal -- median
glomerular filtration rate 90 mL/min (interquartile interval 75.8-96.3) and
median albumin 45 g/L (38-47) -- which matters, because it is the most likely
reason no covariate reached significance.

Infection involved a hip prosthesis in 55% of patients, knee in 30%, shoulder in
10% and ankle in 5%. Coagulase-negative staphylococci accounted for 70.8% of the
24 isolates (Table 1), with *Cutibacterium acnes* at 25% and *Enterococcus
faecalis* at 4.2%; four infections were polymicrobial. Every patient received a
single 1500 mg intravenous dose after a median 11.5 days of prior intravenous
therapy, given as a 30-minute infusion at one hospital and a 2-hour infusion at
the other.

The same information is available programmatically:

```{r population}
str(rxode2::rxode(readModelDb("Benavent_2025_dalbavancin"))$population)
```

# Source trace

Every value in the model file comes from Table 2 of the paper, which is
reproduced in full here. The paper prints no other parameter table.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Quantity,                    ~`Model file`,      ~`Source location`,                                   ~`Published value`,
  "Central volume",             "lvc = log(17.9)",  "Table 2, 'V (L)'",                                   "17.9 L (6.2% RSE) [18.1% shrinkage]; bootstrap 18.0 (16.0-20.3)",
  "Clearance",                  "lcl = log(0.036)", "Table 2, 'CL (L/h)'",                                "0.036 L/h (8.2% RSE) [12.6% shrinkage]; bootstrap 0.037 (0.031-0.043)",
  "IIV on volume",              "etalvc ~ 0.200^2", "Table 2, 'IIV V (SD)'",                              "0.200 SD (27.4% RSE); bootstrap 0.200 (0.061-0.280)",
  "IIV on clearance",           "etalcl ~ 0.290^2", "Table 2, 'IIV CL (SD)'",                             "0.290 SD (20.8% RSE); bootstrap 0.280 (0.150-0.380)",
  "Proportional residual",      "propSd = 0.120",   "Table 2, 'b (proportional)'",                        "0.120 (28.6% RSE); bootstrap 0.110 (0.053-0.170)",
  "Structure",                  "1-cmt IV, kel",    "Results, 'Population pharmacokinetic analysis'",     "'one-compartment model with intravenous infusion and first-order linear elimination'",
  "Error model",                "Cc ~ prop(propSd)","Results, 'Population pharmacokinetic analysis'",     "'the residual error was modeled as proportional'",
  "IIV distribution",           "exp(l* + eta)",    "Methods, 'Structural model building'",               "'log-normally distributed ... IIV was described using an exponential model'",
  "Covariates",                 "none",             "Results, 'Population pharmacokinetic analysis'",     "'The covariate analysis did not result in model improvements'"
) |>
  knitr::kable(caption = "Source trace for every model quantity.")
```

Monolix reports omega on the standard-deviation scale, and Table 2 labels the
rows `IIV V (SD)` / `IIV CL (SD)` with footnote *a* spelling out "SD, standard
deviation", so the model file squares the printed values to variances. This is
the single most consequential reading in the extraction, and the paper leaves no
room for doubt about it.

# Structural check against closed form

For a one-compartment model with no covariates the typical-value profile has an
exact closed form, so this check is deterministic -- it contains no random draw
and the bounds can be tight.

```{r closed-form}
DOSE <- 1500      # mg, single intravenous dose
TINF <- 0.5       # h, 30 min infusion used in the paper's Monte Carlo

typ  <- rxode2::zeroRe(mod)
obs_t <- sort(unique(c(
  seq(0, 4, by = 0.1),        # resolve the end-of-infusion peak
  seq(6, 48, by = 6),
  seq(72, 2520, by = 24),     # out to 15 weeks so AUCinf is well determined
  seq(504, 528, by = 2),      # the three 24 h PK/PD windows (days 21, 27, 35)
  seq(648, 672, by = 2),
  seq(840, 864, by = 2)
)))
ev_typ <- rxode2::et(amt = DOSE, dur = TINF, cmt = "central") |> rxode2::et(obs_t)
sim_typ <- as.data.frame(rxode2::rxSolve(typ, ev_typ))

cl_t  <- 0.036
vc_t  <- 17.9
kel_t <- cl_t / vc_t

# Closed form for a constant-rate infusion of duration TINF, evaluated after
# the infusion has stopped.
cc_closed <- function(t) {
  (DOSE / (cl_t * TINF)) * (1 - exp(-kel_t * TINF)) * exp(-kel_t * (t - TINF))
}

chk <- sim_typ |>
  filter(time >= TINF) |>
  mutate(closed = cc_closed(time), pct = 100 * (Cc - closed) / closed)

c(max_abs_pct_diff = max(abs(chk$pct)),
  cmax_mg_L        = max(sim_typ$Cc),
  half_life_days   = log(2) / kel_t / 24)
```

```{r closed-form-assert}
stopifnot(
  # Deterministic: solver vs closed form, so this is pure numerical error and a
  # tight bound is correct. A mis-transcribed V, CL or infusion duration breaks
  # it immediately.
  max(abs(chk$pct)) < 0.5,
  # Cmax = (Dose / (CL * TINF)) * (1 - exp(-kel * TINF)); 83.8 mg/L.
  abs(max(sim_typ$Cc) - cc_closed(TINF)) < 0.05,
  # Terminal half-life log(2) * V / CL = 344.6 h = 14.4 days.
  abs(log(2) / kel_t / 24 - 14.36) < 0.05
)
```

# Virtual cohort

The paper simulated 1000 profiles. This vignette uses 200, which is the cohort
cap for package vignettes; the resulting Monte Carlo noise on a target-attainment
percentage is a few points, and the assertions below are set outside that noise
(see the thread-count note in "Assumptions and deviations").

```{r cohort}
NSUB <- 200
rxode2::rxSetSeed(20251017)
ev <- rxode2::et(amt = DOSE, dur = TINF, cmt = "central") |>
  rxode2::et(obs_t) |>
  rxode2::et(id = 1:NSUB)

sim <- as.data.frame(rxode2::rxSolve(mod, ev, nSub = NSUB)) |>
  mutate(treatment = "1500 mg IV (30 min infusion)")

stopifnot(nrow(sim) > 0, !anyNA(sim$Cc), all(sim$Cc >= 0),
          length(unique(sim$id)) == NSUB)
```

```{r profile-plot, fig.cap = "Simulated total plasma dalbavancin after a single 1500 mg dose. Median and 5th-95th percentiles over 200 subjects; points are the observed cohort medians reported in Benavent 2025 Results. Replicates the layout of Figure 1 of Benavent 2025 (semi-logarithmic y-axis)."}
obs_pub <- tibble::tibble(
  day  = c(12, 20.5, 28),
  conc = c(58.5, 29.7, 13.3),
  n    = c(4L, 10L, 9L)
)

band <- sim |>
  group_by(day = time / 24) |>
  summarise(lo = quantile(Cc, 0.05), md = median(Cc), hi = quantile(Cc, 0.95),
            .groups = "drop") |>
  # Drop the pre-dose record (zero, so unplottable on a log axis) and truncate
  # to the five weeks Table 3 covers. This is the PLOT window only -- the PKNCA
  # input above keeps every record including time zero.
  filter(day > 0, day <= 35)

ggplot(band, aes(day, md)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.20, fill = "steelblue") +
  geom_line(colour = "steelblue", linewidth = 0.9) +
  geom_point(data = obs_pub, aes(day, conc), colour = "firebrick", size = 2.6) +
  scale_y_log10() +
  labs(x = "Days after the 1500 mg dose", y = "Total dalbavancin (mg/L)",
       title = "Simulated cohort vs the observed medians of Benavent 2025") +
  theme_bw()
```

The three red points are the cohort medians the paper reports at roughly two,
three and four weeks. They are *not* a trajectory: each is a median over a
different, largely non-overlapping handful of patients (n = 4, 10 and 9) sampled
whenever a clinic visit happened to occur. The three-week point sits essentially
on the model median; the two-week and four-week points bracket it in opposite
directions. This is quantified, and treated as a documented deviation rather
than a gate, below.

# PKNCA validation

Benavent 2025 reports no non-compartmental analysis of its own, so there is no
published Cmax / AUC / half-life table to compare against and
`ncaComparisonTable()` is not applicable here. The NCA below instead checks the
simulated cohort against the quantities the model *does* pin down in closed
form: `cl.obs` must recover the population clearance, `vz.obs` the population
volume, and `aucinf.obs` must equal `Dose / CL`.

```{r pknca}
conc_nca <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)

dose_nca <- conc_nca |>
  distinct(id, treatment) |>
  mutate(time = 0, amt = DOSE)

conc_obj <- PKNCA::PKNCAconc(conc_nca, Cc ~ time | treatment + id,
                             concu = "mg/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_nca, amt ~ time | treatment + id,
                             doseu = "mg")

intervals <- data.frame(
  start      = c(0,    504, 648, 840),
  end        = c(Inf,  528, 672, 864),
  cmax       = c(TRUE,  FALSE, FALSE, FALSE),
  tmax       = c(TRUE,  FALSE, FALSE, FALSE),
  auclast    = c(TRUE,  TRUE,  TRUE,  TRUE),
  aucinf.obs = c(TRUE,  FALSE, FALSE, FALSE),
  half.life  = c(TRUE,  FALSE, FALSE, FALSE),
  cl.obs     = c(TRUE,  FALSE, FALSE, FALSE),
  vz.obs     = c(TRUE,  FALSE, FALSE, FALSE)
)

nca <- suppressWarnings(
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
)
nca_res <- as.data.frame(nca)
```

```{r pknca-table, echo = FALSE}
model_implied <- c(
  cmax       = (DOSE / (0.036 * TINF)) * (1 - exp(-(0.036 / 17.9) * TINF)),
  tmax       = TINF,
  aucinf.obs = DOSE / 0.036,
  half.life  = log(2) * 17.9 / 0.036,
  cl.obs     = 0.036,
  vz.obs     = 17.9
)
labels <- c(cmax = "Cmax (mg/L)", tmax = "Tmax (h)",
            aucinf.obs = "AUC0-inf (mg*h/L)", half.life = "t1/2 (h)",
            cl.obs = "CL (L/h)", vz.obs = "Vz (L)")

nca_res |>
  filter(start == 0, PPTESTCD %in% names(model_implied)) |>
  group_by(PPTESTCD) |>
  summarise(sim_median = median(PPORRES), .groups = "drop") |>
  mutate(model = model_implied[PPTESTCD],
         pct   = 100 * (sim_median - model) / model,
         `NCA parameter` = labels[PPTESTCD]) |>
  select(`NCA parameter`,
         `Simulated cohort median` = sim_median,
         `Model-implied typical value` = model,
         `% difference` = pct) |>
  knitr::kable(digits = c(0, 3, 3, 1),
               caption = "Cohort NCA against the closed-form typical values implied by Table 2.")
```

```{r pknca-assert}
nca_med <- nca_res |>
  filter(start == 0) |>
  group_by(PPTESTCD) |>
  summarise(v = median(PPORRES), .groups = "drop")
get <- function(k) {
  x <- nca_med$v[nca_med$PPTESTCD == k]
  if (length(x) != 1L) stop("no unique NCA row for ", k)
  x
}

stopifnot(
  # Cohort medians, so these must admit the draw. Realised across 1 / 2 / 4 /
  # 8 / 16 solver threads: CL -4.5 to +5.0%, Vz -4.1 to +3.1%, AUCinf -4.8 to
  # +4.7%, t1/2 -3.8 to +1.9%. The 12% bound sits outside that spread and
  # still goes red on a mis-transcribed CL or V, which move these by tens of
  # percent.
  abs(100 * (get("cl.obs")     - 0.036)      / 0.036)        < 12,
  abs(100 * (get("vz.obs")     - 17.9)       / 17.9)         < 12,
  abs(100 * (get("aucinf.obs") - DOSE/0.036) / (DOSE/0.036)) < 12,
  abs(100 * (get("half.life")  - log(2)*17.9/0.036) / (log(2)*17.9/0.036)) < 12,
  # Every subject's terminal fit must be clean for a monoexponential model.
  min(nca_res$PPORRES[nca_res$PPTESTCD == "adj.r.squared"]) > 0.99
)
```

# Reproducing Table 3: probability of target attainment

This is the substantive validation. Table 3 gives the probability that
fAUC(0-24h)/MIC stays at or above 50 at three, four and five weeks after the
dose, for four assumed protein-binding values and four MICs -- 48 published
numbers computed from the same model this package now ships.

The unbound exposure is the total AUC over a 24-hour window multiplied by the
free fraction `1 - protein binding`. The AUC comes from PKNCA over the three
interval windows requested above, so no trapezoidal rule is hand-rolled here.

```{r pta}
auc24 <- nca_res |>
  filter(PPTESTCD == "auclast", start > 0) |>
  transmute(id, day = start / 24, auc = PPORRES)

stopifnot(nrow(auc24) == 3L * NSUB)

paper_pta <- tibble::tribble(
  ~pb_pct, ~day, ~`0.03`, ~`0.06`, ~`0.125`, ~`0.25`,
  93L, 21, 100,  100,  100,  99.7,
  93L, 27, 100,  100,  99.6, 97.5,
  93L, 35, 100,  99.5, 97.2, 85.0,
  95L, 21, 100,  100,  100,  99.3,
  95L, 27, 100,  100,  98.8, 92.8,
  95L, 35, 99.6, 98.7, 93.3, 68.9,
  97L, 21, 100,  100,  99.6, 93.4,
  97L, 27, 100,  99.6, 95.5, 66.9,
  97L, 35, 99.3, 96.2, 79.0, 30.6,
  99L, 21, 100,  100,  98.4, 65.9,
  99L, 27, 99.7, 98.2, 87.3, 28.2,
  99L, 35, 97.8, 89.6, 57.0, 5.4
) |>
  pivot_longer(-c(pb_pct, day), names_to = "mic", values_to = "paper") |>
  mutate(mic = as.numeric(mic))

pta_at <- function(fu, mic, day) {
  a <- auc24$auc[auc24$day == day]
  if (length(a) != NSUB) stop("no AUC window for day ", day)
  100 * mean(fu * a / mic >= 50)
}

cmp <- paper_pta |>
  rowwise() |>
  mutate(simulated = pta_at(1 - pb_pct / 100, mic, day)) |>
  ungroup() |>
  mutate(diff = simulated - paper)

stopifnot(nrow(cmp) == 48L, !anyNA(cmp$simulated))
```

```{r pta-table, echo = FALSE}
cmp |>
  mutate(cell = sprintf("%.1f / %.1f", simulated, paper)) |>
  select(pb_pct, day, mic, cell) |>
  pivot_wider(names_from = mic, values_from = cell) |>
  arrange(pb_pct, day) |>
  rename("Protein binding (%)" = pb_pct, "Day post-dose" = day,
         "MIC 0.03" = `0.03`, "MIC 0.06" = `0.06`,
         "MIC 0.125" = `0.125`, "MIC 0.25" = `0.25`) |>
  knitr::kable(caption = "PTA for fAUC(0-24h)/MIC >= 50, simulated / published (Benavent 2025 Table 3).")
```

```{r pta-summary, echo = FALSE}
cmp |>
  group_by(`Protein binding (%)` = pb_pct) |>
  summarise(`Median |difference|` = median(abs(diff)),
            `90th pct |difference|` = quantile(abs(diff), 0.9),
            `Max |difference|` = max(abs(diff)), .groups = "drop") |>
  knitr::kable(digits = 2,
               caption = "Agreement with Table 3, by protein-binding row.")
```

The 93%, 95% and 97% rows reproduce the published values essentially exactly.
The 99% row does not, and the disagreement is enormous -- up to 64 percentage
points. That is not Monte Carlo noise, and the next section shows it is not a
transcription error in this package either.

```{r pta-assert}
low <- cmp |> filter(pb_pct < 99L)

stopifnot(
  # Realised across 1 / 2 / 4 / 8 / 16 solver threads: median |diff|
  # 0.10 / 0.50 / 0.50 / 0.30 / 0.20 and 90th percentile
  # 0.95 / 4.50 / 2.25 / 3.10 / 5.50. The bounds sit outside that spread but
  # still break on a mis-transcribed CL or V, which move whole rows by tens
  # of points.
  median(abs(low$diff)) < 3,
  quantile(abs(low$diff), 0.9) < 12
)
```

# The 99% protein-binding row was computed at 98%

Target attainment here depends on the assumed free fraction and the MIC only
through their ratio `fu / MIC`, because the criterion is
`fu * AUC24 / MIC >= 50`. That makes Table 3 self-checking: any two cells with
the same `fu / MIC` must give the same PTA, whichever row and column they sit
in. They do for the 93%, 95% and 97% rows. They do not for the 99% row, which
reports a *higher* attainment at `fu / MIC = 0.08` (98.4% at day 21) than the
97% row does at the larger ratio 0.12 (93.4%) -- impossible for a monotone
criterion.

Recomputing the fourth row with a free fraction of 2% instead of 1% resolves it:

```{r errata}
row99 <- cmp |>
  filter(pb_pct == 99L) |>
  rowwise() |>
  mutate(`as labelled (fu = 1%)` = simulated,
         `at fu = 2% (98% binding)` = pta_at(0.02, mic, day)) |>
  ungroup()
```

```{r errata-table, echo = FALSE}
row99 |>
  arrange(mic, day) |>
  # MIC is formatted as text: a numeric column would be rounded to the shared
  # `digits` and collapse 0.03 / 0.06 / 0.125 / 0.25 onto 0.0 / 0.1 / 0.2.
  mutate(MIC = format(mic, trim = TRUE, drop0trailing = TRUE)) |>
  select(`Day post-dose` = day, `MIC (mg/L)` = MIC, `Published` = paper,
         `as labelled (fu = 1%)`, `at fu = 2% (98% binding)`) |>
  knitr::kable(digits = 1,
               caption = "The fourth row of Table 3 against the two candidate free fractions.")
```

```{r errata-assert}
d01 <- median(abs(row99$`as labelled (fu = 1%)`       - row99$paper))
d02 <- median(abs(row99$`at fu = 2% (98% binding)`    - row99$paper))
c(`median |diff| at fu = 1%` = d01, `median |diff| at fu = 2%` = d02)

stopifnot(
  # Realised 19.70 / 14.90 / 18.20 / 19.65 / 13.90 (fu = 1%) and
  # 0.50 / 1.90 / 0.85 / 1.10 / 0.65 (fu = 2%) at 1 / 2 / 4 / 8 / 16 solver
  # threads. The two readings are separated by an order of magnitude, far
  # outside the Monte Carlo spread, so both bounds have ample headroom and
  # both can fail.
  d01 > 8,
  d02 < 6
)
```

Every cell of the published fourth row lands within a couple of points of the
2% free fraction, including the two that are most diagnostic because they sit in
the steep part of the curve: 65.9% published against 66.5% simulated at day 21
and MIC 0.25, and 28.2% against 28.0% at day 27. At the labelled 1% free
fraction those same cells simulate at 1.5% and 0.5%.

The paper's own data agree. Its observed median total concentration three weeks
post-dose is 29.7 mg/L; at 99% binding the unbound AUC over the following 24
hours is about 7.1 mg*h/L, which against MIC 0.25 gives a ratio of 28 -- well
below the target of 50, and irreconcilable with a published attainment of 65.9%.
At 98% binding the ratio is 57, which is consistent.

This is a reporting defect in Table 3, not in the model: the affected row is a
downstream simulation product, and the packaged model reproduces the other three
rows to within a few tenths of a percentage point. The conclusion the paper draws
in prose -- that a single 1500 mg dose covers three to four weeks for MICs up to
0.125 mg/L, and up to 0.25 mg/L when binding is 97% or lower -- rests on the rows
that do reproduce and is unaffected.

# Observed concentrations

```{r obs-compare, echo = FALSE}
nearest <- function(d) obs_t[which.min(abs(obs_t - d * 24))]
obs_pub |>
  rowwise() |>
  mutate(`Model median (mg/L)` = median(sim$Cc[sim$time == nearest(day)])) |>
  ungroup() |>
  mutate(`% difference` = 100 * (`Model median (mg/L)` - conc) / conc) |>
  select(`Nominal day` = day, `Patients sampled` = n,
         `Observed median (mg/L)` = conc, `Model median (mg/L)`, `% difference`) |>
  knitr::kable(digits = 1,
               caption = "Published observed cohort medians (Benavent 2025 Results) against the simulated cohort median.")
```

The three-week value agrees closely. The two-week and four-week values differ by
about -20% and +60%, in opposite directions, which no single change to the model
could produce. The mechanism is in how the published numbers were formed: each is
a median over a different small subset of patients (n = 4, 10 and 9) at a nominal
day that is itself an interquartile range (day 12 [8.5-13.5], day 20.5 [18-22],
day 28 [28-29]), sampled when a clinic visit happened rather than on a schedule.
A four-patient median is not an estimate of a population median, and patients
sampled at two weeks are largely not the patients sampled at four weeks. These
values are reported here for transparency and deliberately excluded from the
assertion gate; the quantities that *are* gated are the closed-form structural
check, the cohort NCA, and the 36 reproducible cells of Table 3.

# Assumptions and deviations

* **Total, not unbound, dalbavancin.** The packaged model returns total plasma
  concentration, which is what was measured and fitted. The paper swept four
  theoretical protein-binding values (93, 95, 97 and 99%) rather than adopting
  one, so no single free fraction is packaged; the scaling is applied in this
  vignette at the point of use. This differs from the sibling model
  `Baiardi_2025_dalbavancin`, which does package a free fraction because its
  source paper commits to a single value of 7%.
* **The fourth row of Table 3 is computed at 98% protein binding, not the
  labelled 99%.** Demonstrated above. Treat the published fourth row as
  applying to a free fraction of 2%. The model, and the other 36 published
  cells, are unaffected.
* **Infusion duration is an event-table property.** Patients received either a
  30-minute or a 2-hour infusion depending on the hospital. The paper's Monte
  Carlo used 30 minutes, so this vignette does too. Over a two-week half-life
  the choice is immaterial to exposure beyond the first day.
* **Supplementary material was not retrievable.** Tables S1-S3 and Figures S1-S2
  (file `AAC00773-25-S0001.docx`) could not be fetched from the EuropePMC
  supplementary-files endpoint, the publisher, or PMC. None of them carries a
  model parameter: the complete final model is Table 2 of the main text, which
  is reproduced in full in the source trace above. The consequences are that the
  exact observation count is unknown (18 patients contributing 1-3 samples each,
  so between 18 and 54), the goodness-of-fit and VPC panels could not be
  inspected, and the secondary target-attainment table for the bacteriostatic
  criterion (fAUC/MIC >= 25, Table S3) is not reproduced here.
* **No covariates.** The paper screened age, sex, height, body weight, body mass
  index, glomerular filtration rate, baseline creatinine clearance and same-day
  serum albumin, and retained none. All eight are recorded in the model file's
  `covariatesDataExcluded` so the screen is preserved. The cohort is narrow --
  elderly, normal renal function, normal albumin -- so the absence of a renal or
  albumin effect should be read as "not identifiable in these 18 patients",
  which is how the Discussion reads it too, rather than as evidence of no
  effect. The packaged clearance is about 30% below values reported for healthy
  volunteers (0.036 vs 0.050 L/h), so the model should not be extrapolated to
  younger or renally augmented patients.
* **Cohort size and assertion bounds.** The paper simulated 1000 profiles; this
  vignette simulates 200, the package cap. Every bound in the gates above was
  chosen by measuring the gated statistic at 1, 2, 4, 8 and 16 solver threads
  -- which draw different
  cohorts, because `rxSetSeed()` fixes rxode2's RNG per thread and not across
  thread counts -- and placing the bound outside the observed spread. The
  realised ranges are recorded in comments beside each `stopifnot()` so a later
  reader does not tighten them back onto a single draw.
* **Population metadata.** `n_subjects` is 18, the PK analysis population. The
  demographic summaries (age, sex, renal function, albumin) are reported by the
  paper for all 20 enrolled patients; the paper does not separate the 18 who
  contributed concentrations. Body weight and height are never summarised in the
  main text, so no weight or height distribution is recorded.
