---
title: "Trastuzumab emtansine (T-DM1) preclinical mechanistic DAR PK model (Bender 2014)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trastuzumab emtansine (T-DM1) preclinical mechanistic DAR PK model (Bender 2014)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(ggplot2)
library(tidyr)
```

# Overview

This vignette exercises the **mechanistic model** (Bender 2014 Table II),
which tracks each drug-to-antibody-ratio moiety DAR0, DAR1, …, DAR7
individually. Each moiety has its own three-compartment distribution that
shares the central volume `V1`, peripheral volumes `V2`/`V3`, and
distributional clearances `CLd2`/`CLd3` with every other DAR species. A
first-order deconjugation chain carries DAR_n toward DAR_(n-1) at rates

* `k_7->6 = k_6->5 = k_5->4 = k_4->3 = k_3->2` — a single shared upper-chain
  rate estimated in both species (`k_7->3` in the model file);
* `k_2->1` and `k_1->0` — two separate first-order rate constants.

See the companion `Bender_2014_trastuzumabEmtansine_reduced` model /
vignette for the simpler two-species (T-DM1 + DAR0) lumping (Bender 2014
Table III).

# Scope and species

Cynomolgus monkey estimates populate the default `ini()` block. The rat
parameter set is documented in the model's `population$species_parameters`
metadata and in the model-file comments.

```{r inspect-metadata}
mod <- rxode2::rxode(readModelDb("Bender_2014_trastuzumabEmtansine_mechanistic"))
str(mod$meta$population$species_parameters, max.level = 2)
str(mod$meta$population$dose_products, max.level = 2)
```

# Source trace

Every `ini()` value in `inst/modeldb/specificDrugs/Bender_2014_trastuzumabEmtansine_mechanistic.R`
comes from Bender 2014
([doi:10.1208/s12248-014-9618-3](https://doi.org/10.1208/s12248-014-9618-3))
Table II:

| Parameter     | Cynomolgus | Rat | Notes |
|---------------|------:|------:|-------|
| `CL_TT`       | 17.4 mL/day (CV 24.8%) | 2.42 mL/day (CV 24.0%) | Table II |
| `k_plasma`    | 0.0939 /day | 0.156 /day | Table II |
| `V1`          | 148 mL (CV 11.7%)      | 11.0 mL (CV 18.5%)     | Table II |
| `CLd2`        | 25.5 mL/day            | 49.0 mL/day            | Table II |
| `V2`          | 57.2 mL (CV 46.8%)     | 3.44 mL (CV 49.4%)     | Table II |
| `CLd3`        | 81.2 mL/day            | 12.0 mL/day            | Table II |
| `V3`          | 127 mL                 | 16.7 mL (CV 16.8%)     | Table II |
| `k_7->3`      | 0.341 /day             | 0.543 /day (CV 21.8%)  | Shared rate |
| `k_2->1`      | 0.255 /day             | 0.388 /day             | Table II |
| `k_1->0`      | 0.0939 /day            | 0.114 /day (CV 15.1%)  | Table II |
| Residual      | 15.3% proportional     | 11.1% proportional     | Table II |

# Cynomolgus 30 mg/kg IV T-DM1 DAR 3.1 replication

Bender 2014 Figure 4 (cynomolgus panel) shows TT and per-DAR concentration
profiles after a single 30 mg/kg IV T-DM1 DAR 3.1 dose. Simulating the
mechanistic model requires seeding each `dar*_central` compartment with
the DAR-3.1 dose-product fraction multiplied by the total dose amount.

```{r cyno-sim}
dose_mg <- 120  # 30 mg/kg * 4 kg body weight
dar_fracs <- list(
  dar7_central = 0.02, dar6_central = 0.05,
  dar5_central = 0.10, dar4_central = 0.19,
  dar3_central = 0.26, dar2_central = 0.23,
  dar1_central = 0.13, dar0_central = 0.02
)

events <- rxode2::et()
for (cmt in names(dar_fracs)) {
  events <- rxode2::et(events,
                       amt = dose_mg * dar_fracs[[cmt]],
                       cmt = cmt, time = 0)
}
events <- rxode2::et(events, seq(0.05, 42, length.out = 150), cmt = "Cc")

sim <- rxode2::rxSolve(mod |> rxode2::zeroRe(), events = events)
```

```{r cyno-composite, fig.width = 6.5, fig.height = 4.5}
sim_long <- as.data.frame(sim) |>
  pivot_longer(c(Cc, Ctt), names_to = "analyte", values_to = "conc") |>
  mutate(analyte = recode(analyte,
                          Cc  = "T-DM1 (DAR1-7)",
                          Ctt = "Total trastuzumab"))

ggplot(sim_long, aes(time, conc, colour = analyte)) +
  geom_line(size = 0.9) +
  scale_y_log10(limits = c(0.5, NA)) +
  labs(x = "Time (day)", y = "Concentration (ug/mL)",
       colour = NULL,
       title  = "Cynomolgus 30 mg/kg IV T-DM1 DAR 3.1 (Bender 2014 mechanistic)",
       caption = "Replicates the composite analytes in Bender 2014 Figure 4.")
```

```{r cyno-per-dar, fig.width = 7, fig.height = 4.5}
dar_cols <- paste0("Cdar", 0:7)
sim_dar <- as.data.frame(sim) |>
  select(time, all_of(dar_cols)) |>
  pivot_longer(-time, names_to = "moiety", values_to = "conc") |>
  mutate(moiety = factor(moiety, levels = dar_cols, labels = paste0("DAR", 0:7)))

ggplot(sim_dar, aes(time, conc, colour = moiety)) +
  geom_line() +
  scale_y_log10(limits = c(0.01, NA)) +
  labs(x = "Time (day)", y = "Concentration (ug/mL)",
       colour = "Moiety",
       title  = "Per-DAR concentration profiles (cynomolgus 30 mg/kg DAR 3.1)",
       caption = "Replicates the per-moiety panels in Bender 2014 Figure 4.")
```

The per-DAR profiles reproduce the expected qualitative features from the
paper: DAR7 and DAR6 decline monotonically (no higher species feed them);
DAR3–DAR5 peak early and decay; DAR1 and DAR0 build up over time as they
are fed by deconjugation from the higher moieties.

# Assumptions and deviations

- Cynomolgus body weight fixed at 4 kg to convert paper's mg/kg doses to
  absolute mg; the model treats volumes as absolute (mL in the paper, L in
  the `ini()` block).
- Residual error applied to `Cc` (T-DM1) and `Ctt` (TT) only; per-DAR
  concentrations are exposed as deterministic derived outputs
  (`Cdar0` … `Cdar7`) for VPC-style comparison.
- The model's `k_plasma` parameter is retained for future in-vitro plasma
  stability simulations, even though in the in-vivo ODE the sum
  `CL_in_vivo/V1 + k_plasma` simplifies to `CL_TT/V1`.
