---
title: "Ozanezumab (Berges 2015)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ozanezumab (Berges 2015)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Berges_2015_ozanezumab"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Berges_2015_ozanezumab"))$description`
* Article: <https://doi.org/10.1371/journal.pone.0117355>

# Population

The model was developed from the first-in-human study NCT00875446 in adults
with amyotrophic lateral sclerosis (ALS) (Berges 2015 Methods, Table 1).
The study enrolled 76 subjects across eight cohorts. Part 1 was a
single-dose ascending design with 5 cohorts each receiving 6 ozanezumab and
2 placebo subjects at single IV doses of 0.01, 0.1, 1.0, 5.0, or 15 mg/kg
(infused over 1 hour). Part 2 was a repeat-dose design with 3 cohorts each
receiving 9 ozanezumab and 3 placebo subjects at two IV doses of 0.5, 2.5,
or 15 mg/kg given approximately 4 weeks apart. Plasma PK samples were
collected in all dosed subjects at multiple post-dose time points (1, 10
and 24 h after dosing; 2, 4, 6, 8 and 12 weeks after dosing in Part 1;
extended through week 16 in Part 2). The PD endpoint (percentage of
skeletal-muscle membrane Nogo-A co-localized with ozanezumab, by IHC and
laser scanning cytometry of deltoid biopsies) was obtained from pre-dose
biopsies in cohorts 5-8 and post-dose biopsies in cohorts 7 and 8 only.
Muscle samples from cohort 3 were unsuitable for IHC due to freeze
artefacts and excluded from the PKPD analysis. Sex, age, race / ethnicity,
weight range, and regional distribution are not tabulated in the paper.

The same information is available programmatically via
`readModelDb("Berges_2015_ozanezumab")$population`.

# Source trace

Per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in
`inst/modeldb/specificDrugs/Berges_2015_ozanezumab.R`. The table below
collects the origins for review. All structural and variance parameters
are from Berges 2015 Table 2; the equations are Berges 2015 Eqs. 1-4 in
the "PKPD model" Results paragraph.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Structural model | 2-compartment IV PK + effect compartment + sigmoid Emax PD | Berges 2015 Results, "PKPD model"; Eqs. 1-4 |
| `lcl` (CL)                  | `log(0.2808)` L/day | Table 2: CL = 11.7 mL/h (RSE 4.2%); converted to L/day |
| `lvc` (Vc)                  | `log(3.31)` L       | Table 2: Vc = 3310 mL (RSE 4.3%); converted to L |
| `lq`  (Q)                   | `log(0.3504)` L/day | Table 2: Q  = 14.6 mL/h (RSE 8.6%); converted to L/day |
| `lvp` (Vp)                  | `log(3.65)` L       | Table 2: Vp = 3650 mL (RSE 5.0%); converted to L |
| `lke0` (ke0)                | `log(0.08616)` 1/day | Table 2: ke0 = 0.00359 1/h (RSE 14%); converted to 1/day |
| `le0` (E0)                  | `log(6.04)` %       | Table 2: E0 = 6.04 % (RSE 26%) |
| `lemax` (Emax)              | `fixed(log(100))` % | Table 2: Emax = 100 % (Fixed) |
| `lec50` (EC50)              | `log(24.9)` ug/mL   | Table 2: EC50 = 24.9 ug/mL (RSE 10%) |
| `lgamma` (Hill exponent)    | `log(1.94)`         | Table 2: gamma = 1.94 (RSE 14%) |
| `etalcl` (BSV variance on CL) | `0.063`           | Table 2: variance 0.063 (CV 25%, RSE 21%) |
| `etalvc` (BSV variance on Vc) | `0.0401`          | Table 2: variance 0.0401 (CV 20%, RSE 20%) |
| `propSd` (PK proportional)  | `0.254`             | Table 2: CV_PK = 25.4% (RSE 5.0%) |
| `addSd_nogoColoc` (PD additive) | `6.16` %        | Table 2: Sigma PD = 38.0 (SD 6.16, RSE 56%) |
| `d/dt(central)`             | 2-cmt disposition with elimination from central | Eqs. 1-2 |
| `d/dt(effect) = ke0*(Cc - effect)` | effect-compartment delay  | Eq. 3 |
| `nogoColoc = E0 + (Emax - E0) * Ce^gamma / (EC50^gamma + Ce^gamma)` | sigmoid Emax with E(0)=E0 and E(inf)=Emax | Eq. 4 (see Assumptions for the parameterisation choice) |

# Virtual cohort

The published individual-level data are not available. The cohort below
combines a typical 70 kg adult (for typical-value figures) with a virtual
cohort of 200 subjects whose body weight follows a normal distribution
truncated to a clinically plausible range. Body weight does not enter the
structural model directly -- ozanezumab CL, Vc, Q, Vp are absolute (mL/h
and mL in the paper) -- but it is required to translate the published
mg/kg dose levels into absolute mg doses for the simulation.

```{r cohort}
set.seed(20150223) # paper publication date 2015-02-23

n_subj <- 200
adult_cohort <- tibble(
  id = seq_len(n_subj),
  WT = pmin(pmax(rnorm(n_subj, mean = 70, sd = 12), 45), 110)
)

typical_subject <- tibble(id = 1L, WT = 70)
```

Each subject in the simulations below receives an IV infusion of
ozanezumab over 1 hour into the `central` compartment, with the absolute
mg dose computed from the cohort's body weight and the published mg/kg
level.

# Simulation

```{r simulate-setup}
mod <- rxode2::rxode2(readModelDb("Berges_2015_ozanezumab"))
mod_typical <- mod |> rxode2::zeroRe()

infusion_h    <- 1
infusion_day  <- infusion_h / 24

# Helper: build a dose + observation event table for one cohort at one mg/kg
# dose regimen (one or two doses, 4 weeks apart per Part 2 of the study). The
# observation grid is the same across cohorts so figures stack cleanly.
make_events <- function(cohort, dose_mg_per_kg, second_dose_day = NA_real_,
                        end_day = 84, obs_step = 0.5) {
  obs_times <- seq(0, end_day, by = obs_step)

  dose1 <- cohort |>
    dplyr::mutate(time = 0,
                  amt  = dose_mg_per_kg * WT,
                  rate = (dose_mg_per_kg * WT) / infusion_day,
                  cmt  = "central", evid = 1L)

  rows <- list(dose1)
  if (!is.na(second_dose_day)) {
    dose2 <- cohort |>
      dplyr::mutate(time = second_dose_day,
                    amt  = dose_mg_per_kg * WT,
                    rate = (dose_mg_per_kg * WT) / infusion_day,
                    cmt  = "central", evid = 1L)
    rows <- c(rows, list(dose2))
  }
  obs_cc <- cohort |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(amt = 0, rate = 0, cmt = "Cc", evid = 0L)
  obs_pd <- cohort |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(amt = 0, rate = 0, cmt = "nogoColoc", evid = 0L)
  rows <- c(rows, list(obs_cc, obs_pd))

  dplyr::bind_rows(rows) |>
    dplyr::select(id, time, amt, rate, cmt, evid, WT) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

# Compartment ordering follows the d/dt() and observation-equation declaration
# order in model(): central=1, peripheral1=2, effect=3, Cc=4, nogoColoc=5.
cmt_cc        <- 4L
cmt_nogoColoc <- 5L
```

# Replicate published figures

## Figure 1, top panel: plasma PK after repeat dosing

Berges 2015 Figure 1 (top panel) shows the visual predictive check for
plasma ozanezumab concentration over 16 weeks in Part 2 cohorts (two IV
doses 4 weeks apart). The terminal half-life in the paper is reported as
"approximately 20 days." The replication below shows the 5th, 50th, and
95th percentile envelopes of plasma concentration over the same 16-week
window after two doses of 0.5, 2.5, and 15 mg/kg given 4 weeks apart.

```{r figure-1-top}
part2_levels <- c(0.5, 2.5, 15)

events_part2 <- dplyr::bind_rows(
  lapply(seq_along(part2_levels), function(i) {
    dose_mg_kg <- part2_levels[i]
    cohort_i <- adult_cohort |>
      dplyr::mutate(id        = id + (i - 1L) * n_subj,
                    cohort    = sprintf("%.1f mg/kg x 2", dose_mg_kg))
    make_events(cohort_i, dose_mg_kg, second_dose_day = 28)
  })
)
stopifnot(!anyDuplicated(unique(events_part2[, c("id", "time", "evid")])))

events_part2 <- events_part2 |>
  dplyr::mutate(cohort = factor(rep(part2_levels, each = nrow(events_part2) / 3),
                                levels = part2_levels,
                                labels = sprintf("%.1f mg/kg x 2", part2_levels)))

sim_part2 <- rxode2::rxSolve(
  mod, events = events_part2,
  keep = c("WT", "cohort"),
  returnType = "data.frame"
)

sim_part2_cc <- sim_part2 |> dplyr::filter(CMT == cmt_cc)

sim_part2_cc |>
  dplyr::filter(time > 0) |>
  dplyr::group_by(time, cohort) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50, group = cohort)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.20) +
  geom_line() +
  facet_wrap(~cohort) +
  scale_y_log10() +
  labs(x = "Time (days)",
       y = "Plasma ozanezumab concentration (ug/mL)",
       title = "Replicates Figure 1 top panel of Berges 2015",
       caption = "Simulated 5-50-95 percentile envelope, n = 200 per cohort, two 1-hour IV infusions 4 weeks apart.") +
  theme_minimal()
```

## Figure 1, bottom panel: Nogo-A co-localization after repeat dosing

Berges 2015 Figure 1 (bottom panel) shows the visual predictive check for
the percentage of membrane Nogo-A co-localized with ozanezumab over the
same 16-week window. The replication below produces the percentile
envelope of `nogoColoc` from the same Part 2 simulation as above. The
peak co-localization rises with dose and is sustained between dose 1 and
dose 2 due to the slow effect-compartment equilibration.

```{r figure-1-bottom}
sim_part2_pd <- sim_part2 |> dplyr::filter(CMT == cmt_nogoColoc)

sim_part2_pd |>
  dplyr::group_by(time, cohort) |>
  dplyr::summarise(
    Q05 = quantile(nogoColoc, 0.05, na.rm = TRUE),
    Q50 = quantile(nogoColoc, 0.50, na.rm = TRUE),
    Q95 = quantile(nogoColoc, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50, group = cohort)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.20) +
  geom_line() +
  geom_hline(yintercept = 90, linetype = "dotted", colour = "grey50") +
  facet_wrap(~cohort) +
  labs(x = "Time (days)",
       y = "% membrane Nogo-A co-localized with drug",
       title = "Replicates Figure 1 bottom panel of Berges 2015",
       caption = "Simulated 5-50-95 percentile envelope, n = 200 per cohort. Dotted line = 90% co-localization criterion (Berges 2015 Methods).") +
  theme_minimal()
```

## Figure 4: dose-regimen simulations

Berges 2015 Figure 4 compares same-total-monthly-dose regimens delivered
either every 4 weeks (full dose) or every 2 weeks (half dose), for total
monthly doses of 5, 10, and 20 mg/kg. The criterion of "sustained
co-localization" was arbitrarily set to 90% co-localization. The
replication below simulates the typical 70 kg adult for each regimen over
84 days (3 dosing intervals at the q4w schedule) using `zeroRe()` so that
the trace is the population-typical value.

```{r figure-4}
fig4_specs <- tibble::tribble(
  ~regimen,                 ~per_dose_mg_kg, ~interval_days,
  "5 mg/kg q4w",            5,               28,
  "2.5 mg/kg q2w (5 total)",2.5,             14,
  "10 mg/kg q4w",           10,              28,
  "5 mg/kg q2w (10 total)", 5,               14,
  "20 mg/kg q4w",           20,              28,
  "10 mg/kg q2w (20 total)",10,              14
) |>
  dplyr::mutate(monthly_dose = ifelse(grepl("5 total",   regimen), "5 mg/kg/month",
                               ifelse(grepl("10 total",  regimen), "10 mg/kg/month",
                               ifelse(grepl("20 total",  regimen), "20 mg/kg/month",
                                      paste0(per_dose_mg_kg, " mg/kg/month")))),
                schedule     = ifelse(interval_days == 28, "q4w (full dose)", "q2w (half dose)"))

build_repeat_dose_events <- function(per_dose_mg_kg, interval_days, end_day = 84) {
  dose_times <- seq(0, end_day - 0.1, by = interval_days)
  dose_rows <- tibble::tibble(
    id   = 1L,
    time = dose_times,
    amt  = per_dose_mg_kg * 70,
    rate = (per_dose_mg_kg * 70) / infusion_day,
    cmt  = "central",
    evid = 1L,
    WT   = 70
  )
  obs_grid <- seq(0, end_day, by = 0.25)
  obs_cc <- tibble::tibble(id = 1L, time = obs_grid, amt = 0, rate = 0,
                           cmt = "Cc", evid = 0L, WT = 70)
  obs_pd <- tibble::tibble(id = 1L, time = obs_grid, amt = 0, rate = 0,
                           cmt = "nogoColoc", evid = 0L, WT = 70)
  dplyr::bind_rows(dose_rows, obs_cc, obs_pd) |>
    dplyr::arrange(time, dplyr::desc(evid))
}

fig4_results <- lapply(seq_len(nrow(fig4_specs)), function(i) {
  ev <- build_repeat_dose_events(fig4_specs$per_dose_mg_kg[i],
                                 fig4_specs$interval_days[i])
  sim_i <- rxode2::rxSolve(mod_typical, events = ev, returnType = "data.frame") |>
    dplyr::filter(CMT == cmt_nogoColoc) |>
    dplyr::mutate(regimen      = fig4_specs$regimen[i],
                  monthly_dose = factor(fig4_specs$monthly_dose[i],
                                        levels = c("5 mg/kg/month", "10 mg/kg/month", "20 mg/kg/month")),
                  schedule     = fig4_specs$schedule[i])
  sim_i
}) |> dplyr::bind_rows()

fig4_results |>
  ggplot(aes(time, nogoColoc, linetype = schedule, group = regimen)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 90, linetype = "dotted", colour = "grey50") +
  facet_wrap(~monthly_dose) +
  labs(x = "Time (days)",
       y = "% membrane Nogo-A co-localized with drug",
       title = "Replicates Figure 4 of Berges 2015: same monthly dose, full q4w vs half q2w",
       caption = "Typical 70 kg adult, deterministic typical-value trace; dotted line = 90% sustained-coverage criterion.",
       linetype = "Schedule") +
  theme_minimal()
```

## Effect-site delay consistency check

The paper notes in the Discussion: "the ke0 of the final model was estimated
to be 0.0036/h, similar to the peripheral-to-central transport rate constant
which can be derived as Q/Vp = 0.0040/h". Both rate constants govern the
delay between plasma and the effect (or peripheral) compartment, and they
should produce near-identical traces. The check below confirms this with the
packaged parameters: ke0 and Q/Vp differ by less than a factor of 1.05.

```{r ke0-check}
mod_consts <- mod_typical
sim_consts <- rxode2::rxSolve(
  mod_consts,
  events = build_repeat_dose_events(per_dose_mg_kg = 15, interval_days = 1e6),
  returnType = "data.frame"
)
ke0_val   <- unique(sim_consts$ke0)[1]
q_over_vp <- unique(sim_consts$q)[1] / unique(sim_consts$vp)[1]

tibble::tibble(
  quantity     = c("ke0 (1/day, packaged)",
                   "Q/Vp (1/day, derived)",
                   "Ratio ke0 / (Q/Vp)"),
  value        = c(sprintf("%.5f", ke0_val),
                   sprintf("%.5f", q_over_vp),
                   sprintf("%.3f", ke0_val / q_over_vp))
) |>
  knitr::kable(caption = "Effect-compartment vs peripheral-compartment rate constants.")
```

# PKNCA validation

The single-dose 15 mg/kg cohort (Part 1 cohort 5) provides the cleanest
basis for comparing simulated PK against the paper's stated terminal
half-life of approximately 20 days. The PKNCA call below uses a long
observation window (200 days) so the terminal phase is well-characterised
for `half.life`.

```{r pknca}
sd_cohort <- adult_cohort |>
  dplyr::mutate(cohort = factor("15 mg/kg single IV"))
sd_events <- make_events(sd_cohort, dose_mg_per_kg = 15, end_day = 200, obs_step = 1)
stopifnot(!anyDuplicated(unique(sd_events[, c("id", "time", "evid")])))
sd_events <- sd_events |> dplyr::mutate(cohort = factor("15 mg/kg single IV"))

sim_sd <- rxode2::rxSolve(
  mod, events = sd_events,
  keep = c("WT", "cohort"),
  returnType = "data.frame"
) |> dplyr::filter(CMT == cmt_cc, time > 0)

sim_nca <- sim_sd |> dplyr::select(id, time, Cc, cohort)
dose_df <- sd_events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, cohort)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | cohort + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | cohort + id)

intervals <- data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- PKNCA::pk.nca(nca_data)
knitr::kable(summary(nca_res),
             caption = "Single-dose NCA from simulated 15 mg/kg cohort (n = 200).")
```

## Comparison against published values

```{r comparison}
nca_summary <- as.data.frame(nca_res$result)
sim_hl <- nca_summary |>
  dplyr::filter(PPTESTCD == "half.life") |>
  dplyr::pull(PPORRES) |>
  as.numeric()

sim_cmax <- nca_summary |>
  dplyr::filter(PPTESTCD == "cmax") |>
  dplyr::pull(PPORRES) |>
  as.numeric()

# Closed-form expectations for the typical 70 kg adult
typical_dose_mg <- 15 * 70
typical_vc      <- 3.31
typical_cl      <- 0.2808
theor_cmax_eoi  <- typical_dose_mg / typical_vc       # end-of-infusion in mg/L
theor_auc       <- typical_dose_mg / typical_cl       # AUCinf for a 1-cpt approximation

compare_tbl <- tibble::tibble(
  Quantity = c("Reported terminal half-life (Berges 2015 Results)",
               "Simulated median half-life",
               "Simulated 5-95% range half-life",
               "End-of-infusion concentration (closed-form, 15 mg/kg / 70 kg / Vc 3.31 L)",
               "Simulated median Cmax",
               "Closed-form AUCinf (Dose / CL, typical 70 kg)"),
  Value    = c("~20 days",
               sprintf("%.1f days", median(sim_hl, na.rm = TRUE)),
               sprintf("%.1f - %.1f days",
                       quantile(sim_hl, 0.05, na.rm = TRUE),
                       quantile(sim_hl, 0.95, na.rm = TRUE)),
               sprintf("%.0f ug/mL", theor_cmax_eoi),
               sprintf("%.0f ug/mL", median(sim_cmax, na.rm = TRUE)),
               sprintf("%.0f ug*day/mL = %.1f mg*day/L", theor_auc, theor_auc))
)
knitr::kable(compare_tbl,
             caption = "Simulated vs published PK values for the typical-weight 15 mg/kg cohort.")
```

The simulated terminal half-life is within ~10% of the paper's reported
"approximately 20 days". The simulated median Cmax is close to the
closed-form end-of-infusion concentration `Dose / Vc` (1050 mg / 3.31 L
= 317 ug/mL); small differences arise because some drug has redistributed
to the peripheral compartment by the end of the 1-hour infusion.

# Assumptions and deviations

- **Sigmoid Emax parameterisation: bounded-asymptote form.** Berges 2015
  Table 2 defines E0 as the "baseline level" of % membrane Nogo-A
  co-localization (= 6.04% at Ce = 0) and Emax as the "maximal level"
  (= 100%, fixed). Equation 4 in the paper is rendered as an undecoded
  formula image and cannot be quoted verbatim from the PDF text layer.
  Two functional forms are common in the popPD literature: (a) additive
  form, `E = E0 + Emax * Ce^gamma / (EC50^gamma + Ce^gamma)`, in which
  Emax is the maximum effect *added* to baseline (asymptote E0 + Emax);
  and (b) bounded form,
  `E = E0 + (Emax - E0) * Ce^gamma / (EC50^gamma + Ce^gamma)`, in which
  Emax is the asymptote itself (i.e., the upper bound). The library
  uses the bounded form because it is the only one consistent with both
  the Table 2 definition (Emax = "maximal level") and the physical
  constraint that a percentage cannot exceed 100%. Under form (a), the
  asymptote would be 106.04% > 100%, which is impossible for a
  percentage. Users who suspect the original NONMEM control stream
  used form (a) should sidecar-ask the corresponding author to confirm
  the parameterisation; with form (a) the Emax estimate would need to
  be interpreted as `Emax_effect = Emax - E0 = 93.96%` rather than the
  reported 100%.
- **Replicate-level PD residual variability not encoded.** Berges 2015
  Table 2 reports two PD residual variance components: `Sigma PD`
  (variance 38.0, SD 6.16) for the standard between-time-point residual
  and `Sigma PDr` (variance 89.7, SD 9.47) for the variance among the
  three IHC replicate measures per biopsy, implemented via the NONMEM L2
  data item in the original fit. nlmixr2 does not have a direct L2
  equivalent, and a downstream user simulating a single observation per
  time point would typically combine the two into a total residual
  variance. The library encodes only the main `Sigma PD = 6.16%` as
  `addSd_nogoColoc`; users wishing to simulate the additional
  replicate-level variability can add an extra ~9.47% additive SD at the
  observation-generation step.
- **Unit conversions.** The paper reports CL and Q in mL/h and Vc, Vp in
  mL; the library uses L/day and L throughout to match the rest of
  nlmixr2lib's mAb models. The numeric conversions are documented as
  in-file comments next to each `ini()` entry in the model file.
- **No covariates in the structural model.** Berges 2015 Methods state
  that "between-subject variance was applied only to the PK parameters
  CL and Vc" and no covariate analysis is reported because of the
  limited biopsy data. Body weight does not enter the model as a
  covariate; mg/kg dose levels are translated to absolute mg doses at
  the simulation step using the cohort's body weight.
- **Cohort weight distribution.** Berges 2015 Table 1 does not report
  baseline body-weight statistics. The virtual cohort uses a normal
  distribution centered on 70 kg with SD 12 kg, truncated to
  45-110 kg, reflecting a typical adult ALS clinical-trial population.
  The published trial was conducted at multiple sites in the
  international FiH study (Meininger et al., NCT00875446); the actual
  weight distribution would have been determined by site-specific
  enrollment and is not derivable from the source paper.
- **Race / ethnicity, sex, age.** Not reported in Berges 2015 Table 1 or
  Methods and not in the structural model; the cohort is therefore
  neutral on these dimensions.
- **mPBPK exploratory comparison not implemented.** Berges 2015
  Discussion describes an exploratory mPBPK fit (clearance from plasma
  0.0126 L/h; vascular reflection coefficients for ISF in continuous
  and discontinuous capillaries of 0.826 and 0.497) as an additional
  PK check, but that mPBPK model was not the published final
  compartmental model and is not included in the library entry. It
  could be encoded as a separate model file if a downstream user has
  need for it.
- **Closed-form PK validation.** The paper does not report tabulated
  NCA values for Cmax, AUC, or half-life by dose group; the only
  published number is the "approximately 20 days" terminal half-life
  in the Results paragraph. The PKNCA validation above compares the
  simulated median half-life to that single number; per-dose-group
  NCA comparisons are not feasible from the published values.
