---
title: "Levosimendan (Bertin 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Levosimendan (Bertin 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Bertin_2026_levosimendan"))$reference`
* Description: `r rxode2::rxode(readModelDb("Bertin_2026_levosimendan"))$description`
* Article: <https://doi.org/10.1007/s40262-025-01591-4>
* Electronic Supplementary Material (back-transformation-constant derivation,
  covariate-analysis tables, goodness-of-fit and pvc-VPC plots, the additional
  dosing-scenario simulations, and -- decisively for this extraction -- the full
  final NONMEM control stream) is distributed with the same DOI as
  `40262_2025_1591_MOESM1_ESM.pdf`.

Levosimendan is a calcium-sensitising inotrope and vasodilator used in acute
heart failure and, in particular, to facilitate weaning from veno-arterial
extracorporeal membrane oxygenation (VA-ECMO). Its clinical profile is
dominated by its metabolites rather than by the parent: levosimendan itself
has an elimination half-life of about one hour, but roughly 5% of the dose is
reduced by gut microbiota to OR-1855, which polymorphic N-acetyltransferase-2
then acetylates to OR-1896. OR-1896 reproduces levosimendan's haemodynamic
effects and has a half-life of 70-80 hours, so it -- not the parent -- is
responsible for the inotropic support that persists for up to a week after an
infusion stops.

This is a bicentric prospective study of 21 critically ill patients on
VA-ECMO, and the first popPK model to describe all three species together in
adults *and* neonates/infants. Its central finding is that in patients aged one
year or younger the acetylation of OR-1855 to OR-1896 is 3.7-fold slower than
in adults, so the sustained post-infusion effect that clinicians rely on may
be substantially reduced or absent in that group.

## Why the supplement is load-bearing here

Three facts needed to reproduce this model are **not** in the main paper, and
come only from the control stream in the ESM. They are called out here because
an extraction built from Table 2 alone would be quietly wrong:

1. **`ktransit` is 0.0127 1/h, not 0.01.** Table 2 rounds it to one
   significant figure. The transit rate constant sets the entire timescale on
   which both metabolites appear, so a 27% error here propagates to every
   metabolite concentration.
2. **The metabolite-formation flux is carved *out of* clearance, not added to
   it.** The control stream computes `K10 = CL/V1 - K13`, so levosimendan
   still leaves the central compartment at a total rate of `CL/V1`. Encoding
   the transit arm additively is the natural mistake and is checked against
   the published half-lives below.
3. **"Childhood" means age 1 year or younger.** The paper names the covariate
   only as "childhood" or "neonates/infants vs adults" and never gives a
   cut-off; `IF(AGE.LE.1) Q1=1` in the `$PK` block does.

A fourth discrepancy is internal to the paper: Table 2's "Final model
estimate" column prints the OR-1896-to-OR-1855 back-conversion constant as
`0.01 FIX`, while its own bootstrap column, the Methods text and the control
stream all give 0.012. The model uses 0.012.

# Population

```{r population, echo = FALSE}
pop <- rxode2::rxode(readModelDb("Bertin_2026_levosimendan"))$population
tibble::tibble(Field = names(pop), Value = vapply(pop, function(x) paste(as.character(x), collapse = "; "), character(1))) |>
  knitr::kable(caption = "Study population metadata (Bertin 2026 Table 1 and Sect. 2.1/3.1).")
```

Twenty-one patients -- 15 adults, 3 neonates and 3 infants -- contributed 155
plasma samples. Adults started at 0.05 ug/kg/min and were escalated within
4 hours to a maintenance rate of 0.1, 0.15 or 0.2 ug/kg/min for about 24 hours;
the neonates and infants all received 0.1 ug/kg/min for 48 hours. All 21 were
on veno-arterial ECMO. Sampling stopped 24 hours after the end of the infusion,
which is why the metabolites' elimination was never observed and their
elimination rate constants had to be fixed from the literature.

# Source trace

```{r source_trace, echo = FALSE}
tibble::tribble(
  ~Quantity, ~Value, ~`Source location`,
  "CL",                  "13.9 L/h",   "ESM Suppl. 10 $THETA 1; Table 2 'CL (L/h) 14 (21%)'",
  "theta_BW on CL",      "0.75 FIX",   "ESM Suppl. 10 $THETA 2; Table 2; Sect. 3.2 (allometric theory)",
  "V1",                  "15.9 L",     "ESM Suppl. 10 $THETA 3; Table 2 'V1 (L) 16 (26%)'",
  "theta_BW on V1",      "0.574",      "ESM Suppl. 10 $THETA 4; Table 2 '0.57 (32%)'",
  "Q",                   "0.501 L/h",  "ESM Suppl. 10 $THETA 5; Table 2 'Q (L/h) 0.50 (36%)'",
  "V2",                  "5.75 L",     "ESM Suppl. 10 $THETA 6; Table 2 'V2 (L) 5.8 (44%)'",
  "ktransit",            "0.0127 1/h", "ESM Suppl. 10 $THETA 7 (Table 2 rounds to 0.01)",
  "keM1",                "0.01 1/h FIX", "ESM Suppl. 10 $THETA 8; Sect. 2.3.1 (t-half approx. 70 h)",
  "theta_BW on keM1",    "-0.611",     "ESM Suppl. 10 $THETA 13; Table 2 '- 0.61 (22%)'",
  "kM2 (adults)",        "0.0722 1/h", "ESM Suppl. 10 $THETA 9; Table 2 '0.07 (44%)'",
  "theta_child on kM2",  "-0.732",     "ESM Suppl. 10 $THETA 12; Table 2 '- 0.73 (30%)'",
  "keM2",                "0.01 1/h FIX", "ESM Suppl. 10 $THETA 10; Sect. 2.3.1",
  "kM2-M1",              "0.012 1/h FIX", "ESM Suppl. 10 $THETA 11; ESM Suppl. 1 derivation; Sect. 2.3.1",
  "Reference weight",    "70 kg",      "Sect. 3.2 final-model equations; $PK 'MWT = 70'",
  "Childhood cut-off",   "age <= 1 year", "ESM Suppl. 10 $PK 'IF(AGE.LE.1) Q1=1'",
  "omega CL / V1 / V2",  "0.0998 / 0.235 / 0.684", "ESM Suppl. 10 $OMEGA 1, 2, 4 (variances)",
  "omega ktransit / kM2","0.128 / 0.599", "ESM Suppl. 10 $OMEGA 5, 7 (variances)",
  "sigma levo / M1 / M2","0.0952 / 0.135 / 0.0916", "ESM Suppl. 10 $SIGMA 1-3 (variances)",
  "ODE system",          "5 states",   "ESM Suppl. 10 $MODEL and $DES; Fig. 2 schematic",
  "V3 = V4 = V1",        "assumption", "Sect. 2.3.1 (identifiability); $PK 'V4 = V1', 'V5 = V1'"
) |>
  knitr::kable(caption = "Provenance of every model equation and ini() value.")
```

# Molecular weights recovered from the control stream

The source analysis works entirely in molar units, so reproducing Table 3 -- which is reported in ng/mL -- needs the three molecular weights. The paper never
prints them, but they are recoverable exactly: the assay LLOQ was 0.1 ng/mL for
all three analytes (Sect. 2.2), and the `$ERROR` block carries the natural log
of that same limit expressed in the model's molar units.

```{r mw_recovery}
ln_loq <- c(levosimendan = -7.938, or1855 = -7.617, or1896 = -7.805)
mw <- 0.1 / exp(ln_loq)

# Independent cross-check: the ESM pvc-VPC captions print the same limits
# rounded to two significant figures, in nmol/mL.
loq_molar_published <- c(levosimendan = 0.00036, or1855 = 0.00049, or1896 = 0.00041)

tibble::tibble(
  Analyte = names(mw),
  `MW recovered (g/mol)` = round(mw, 2),
  `Formula weight (g/mol)` = c(280.28, 203.24, 245.28),
  `LOQ implied (nmol/mL)` = signif(exp(ln_loq), 2),
  `LOQ in ESM captions` = loq_molar_published
) |>
  knitr::kable(caption = "Molecular weights recovered from the $ERROR block's ln(LOQ) constants.")

# The recovered weights must match the compounds' formula weights. The
# tolerance is set by the 3-decimal rounding of the printed ln(LOQ), which is
# worth about 0.05% on the recovered mass.
stopifnot(
  max(abs(mw - c(280.28, 203.24, 245.28)) / c(280.28, 203.24, 245.28)) < 0.001,
  all(abs(signif(exp(ln_loq), 2) - loq_molar_published) < 1e-9)
)
```

Levosimendan is C14H12N6O2 (280.28), OR-1855 its amino reduction product
C11H13N3O (203.24), and OR-1896 the acetylated form C13H15N3O2 (245.28). The
recovered weights agree to within the rounding of the printed constants, which
confirms both the molar parameterisation and the identity of each compartment.

```{r mw_constants, echo = FALSE}
MW_LEVO <- 280.28
MW_M1 <- 203.24
MW_M2 <- 245.28
```

# Structural verification against the published closed-form results

These checks use the **typical-value** model (`zeroRe()`), so they are exact
and deterministic -- no cohort is drawn and no random-number stream is
involved. They are the tight gates of this vignette; the cohort comparison
later is necessarily looser.

```{r structural}
mod <- readModelDb("Bertin_2026_levosimendan")

# Every number on the "Simulated" side of the table below is taken from the
# model file's own ini() block, never retyped from the paper. That direction
# matters: the "Published" column holds the paper's printed figures, so a
# mis-transcribed parameter in the model file makes a row go red. A gate that
# hardcoded both sides would only be checking the paper against itself.
th <- rxode2::rxode(mod)$theta

# Closed-form two-compartment disposition of the parent. The point of this
# check is the `kel <- cl/vc - ktr` coupling: if the transit arm were added to
# clearance instead of carved out of it, t-half-alpha for a 70-kg adult would
# come out at 0.75 h rather than the published 0.76 h.
disposition <- function(WT, th) {
  cl <- exp(th[["lcl"]]) * (WT / 70)^th[["e_wt_cl"]]
  vc <- exp(th[["lvc"]]) * (WT / 70)^th[["e_wt_vc"]]
  q <- exp(th[["lq"]])
  vp <- exp(th[["lvp"]])
  k10 <- cl / vc
  k12 <- q / vc
  k21 <- q / vp
  s <- k10 + k12 + k21
  d <- sqrt(s^2 - 4 * k10 * k21)
  c(
    t_half_alpha = log(2) / ((s + d) / 2),
    t_half_beta = log(2) / ((s - d) / 2),
    cl = cl,
    cl_per_kg_mL_min = cl / WT * 1000 / 60,
    vss_L_per_kg = (vc + vp) / WT
  )
}
adult <- disposition(70, th)
child <- disposition(4, th)

# Metabolite-formation rate constant, adults and the age <= 1 group. The
# childhood term is the categorical form kM2 * (1 + child * theta), so the
# paediatric factor is 1 + e_child_kmet_or1896.
km2_adult <- exp(th[["lkmet_or1896"]])
child_factor <- 1 + th[["e_child_kmet_or1896"]]

# `half_ulp` is half of the last printed digit of each published figure, i.e.
# the largest discrepancy that rounding alone can explain. Asserting against
# it per row is much stricter than a blanket tolerance: CL at 70 kg is printed
# as 13.9 and so must agree to 0.36%, while CL at 4 kg is printed as 1.6 and
# can only be pinned to 3.1%.
published <- tibble::tribble(
  ~Quantity,                              ~Simulated,                  ~Published, ~half_ulp, ~`Source`,
  "t-half alpha, 70-kg adult (h)",        adult[["t_half_alpha"]],     0.76,       0.005,     "Sect. 3.2",
  "t-half beta, 70-kg adult (h)",         adult[["t_half_beta"]],      8.3,        0.05,      "Sect. 3.2",
  "t-half alpha, 4-kg child (h)",         child[["t_half_alpha"]],     0.97,       0.005,     "Sect. 3.2",
  "t-half beta, 4-kg child (h)",          child[["t_half_beta"]],      10.7,       0.05,      "Sect. 3.2",
  "CL, 70 kg (L/h)",                      adult[["cl"]],               13.9,       0.05,      "Sect. 3.2",
  "CL, 100 kg (L/h)",                     disposition(100, th)[["cl"]], 18.2,      0.05,      "Sect. 3.2",
  "CL, 4 kg (L/h)",                       child[["cl"]],               1.6,        0.05,      "Sect. 3.2",
  "CL/kg, 4 kg (mL/min/kg)",              child[["cl_per_kg_mL_min"]], 6.77,       0.005,     "Sect. 4",
  "Vss, 4 kg (L/kg)",                     child[["vss_L_per_kg"]],     2.21,       0.005,     "Sect. 4",
  "kM2, adults (1/h)",                    km2_adult,                   0.072,      0.0005,    "Sect. 3.2",
  "kM2, neonates/infants (1/h)",          km2_adult * child_factor,    0.019,      0.0005,    "Sect. 3.2",
  "kM2 adult:child ratio",                1 / child_factor,            3.7,        0.05,      "Sect. 3.2"
) |>
  dplyr::mutate(
    `Diff (%)` = 100 * (Simulated - Published) / Published,
    `Rounding allows (%)` = 100 * half_ulp / Published
  )

published |>
  dplyr::mutate(Simulated = signif(Simulated, 4)) |>
  dplyr::select(-half_ulp) |>
  knitr::kable(caption = "Closed-form model quantities versus the values Bertin 2026 reports.", digits = 3)

# Each of these is a deterministic function of the ini() values, so the only
# thing that can move them is a mis-transcribed parameter. Every row must agree
# with the paper to within the precision at which the paper printed it -- i.e.
# the model reproduces each published figure exactly, as far as can be told
# from the digits given.
stopifnot(all(abs(published$Simulated - published$Published) <= published$half_ulp))
```

All twelve reproduce to within the precision at which the paper printed them.
Because the four half-lives are jointly determined by
CL, V1, Q, V2 *and* the `kel <- cl/vc - ktr` coupling, and because CL/kg and
Vss pin the two body-weight exponents independently, this single table
validates the entire parent-disposition layer and both allometric terms.
And because the simulated column is computed from the model file's `ini()`
values rather than from numbers retyped out of the paper, the table fails if
any one of `lcl`, `lvc`, `lq`, `lvp`, `e_wt_cl`, `e_wt_vc`, `lkmet_or1896` or
`e_child_kmet_or1896` is mis-transcribed.

# Dosing scenarios

The paper simulates four scenarios (Sect. 2.3.3, Table 3). Doses are infusion
rates in ug/kg/min, which must be converted to the model's umol/h.

```{r scenarios}
# ug/kg/min -> umol/h:  rate * WT [kg] * 60 [min/h] / MW [ug/umol]
to_umol_per_h <- function(ug_kg_min, WT) ug_kg_min * WT * 60 / MW_LEVO

scenarios <- tibble::tribble(
  ~scenario,     ~WT, ~AGE,  ~rates,              ~durations,   ~label,
  "Scenario 1",  70,  62,    c(0.05, 0.1),        c(1, 23),     "Adult 70 kg: 0.05 for 1 h, then 0.1 ug/kg/min for 23 h",
  "Scenario 2",  70,  62,    c(0.05, 0.1, 0.2),   c(1, 3, 20),  "Adult 70 kg: escalated to 0.2 ug/kg/min at 4 h",
  "Scenario 3",  4,   0.07,  0.1,                 48,           "Neonate/infant 4 kg: 0.1 ug/kg/min for 48 h",
  "Scenario 4",  4,   0.07,  0.2,                 48,           "Neonate/infant 4 kg: 0.2 ug/kg/min for 48 h"
)

knitr::kable(
  scenarios |> dplyr::select(scenario, WT, AGE, label) |>
    dplyr::rename("Scenario" = scenario, "Weight (kg)" = WT, "Age (years)" = AGE, "Regimen" = label),
  caption = "The four dosing scenarios of Bertin 2026 Table 3."
)
```

`AGE` enters only through the `AGE <= 1` childhood indicator, so the adult
value (the cohort median, 62 years) and the neonate value (24 days = 0.07
years, the cohort median) simply place each arm on the correct side of the
cut-off.

```{r event_builder}
# Event tables are built as data frames, not rxEt objects, so that the
# covariate columns survive. Observation rows carry cmt = "central" (a real
# ODE state) plus dvid = 1 to nominate an endpoint, which is required because
# the model declares three. All three observable columns are returned on every
# observation row regardless of which endpoint the row nominates, so one set
# of rows serves all three analytes.
build_events <- function(WT, AGE, rates, durations, ids, tmax = 200, by = 0.5) {
  r <- to_umol_per_h(rates, WT)
  starts <- c(0, cumsum(durations)[-length(durations)])
  do.call(rbind, lapply(ids, function(i) {
    rbind(
      data.frame(id = i, time = starts, amt = r * durations, rate = r,
                 evid = 1L, cmt = "central", dvid = NA_integer_),
      data.frame(id = i, time = seq(0, tmax, by = by), amt = NA_real_, rate = NA_real_,
                 evid = 0L, cmt = "central", dvid = 1L)
    )
  })) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    dplyr::mutate(WT = WT, AGE = AGE)
}
```

# Typical-value profiles (replicates Figure 3)

```{r typical_profiles, fig.height = 6}
mod_typ <- rxode2::zeroRe(mod)

typical <- dplyr::bind_rows(lapply(seq_len(nrow(scenarios)), function(i) {
  s <- scenarios[i, ]
  ev <- build_events(s$WT, s$AGE, s$rates[[1]], s$durations[[1]], ids = 1L)
  rxode2::rxSolve(mod_typ, ev, returnType = "data.frame") |>
    dplyr::mutate(scenario = s$scenario)
}))

profiles <- typical |>
  dplyr::transmute(
    scenario, time,
    Levosimendan = Cc * MW_LEVO,
    `OR-1855` = Cc_or1855 * MW_M1,
    `OR-1896` = Cc_or1896 * MW_M2
  ) |>
  tidyr::pivot_longer(c(Levosimendan, `OR-1855`, `OR-1896`),
                      names_to = "Analyte", values_to = "conc")

ggplot(profiles, aes(time, conc, colour = scenario)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~Analyte, ncol = 1, scales = "free_y") +
  labs(
    x = "Time (h)", y = "Concentration (ng/mL)", colour = NULL,
    title = "Typical-value concentration-time profiles",
    subtitle = "Replicates the shape of Bertin 2026 Figure 3 (scenarios 1 and 3) and ESM Suppl. 6 (scenarios 2 and 4)"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

The three panels reproduce the qualitative behaviour the paper describes: the
parent reaches a plateau within a few hours and washes out quickly once the
infusion stops, while both metabolites keep rising long after the infusion has
ended -- OR-1896 does not peak until roughly day 5.

# Virtual cohort

```{r cohort}
rxode2::rxSetSeed(20260911)
N_PER_ARM <- 200

cohort <- dplyr::bind_rows(lapply(seq_len(nrow(scenarios)), function(i) {
  s <- scenarios[i, ]
  ev <- build_events(s$WT, s$AGE, s$rates[[1]], s$durations[[1]], ids = seq_len(N_PER_ARM))
  rxode2::rxSolve(mod, ev, returnType = "data.frame") |>
    dplyr::mutate(scenario = s$scenario)
}))

# Convert to the paper's mass units once, here.
cohort <- cohort |>
  dplyr::mutate(
    levosimendan = Cc * MW_LEVO,
    or1855 = Cc_or1855 * MW_M1,
    or1896 = Cc_or1896 * MW_M2
  )

dplyr::count(cohort, scenario, name = "rows") |>
  dplyr::mutate(subjects = N_PER_ARM) |>
  dplyr::rename("Scenario" = scenario, "Rows" = rows, "Subjects" = subjects) |>
  knitr::kable(caption = "Simulated cohort size (200 subjects per arm).")
```

The paper simulated 1000 subjects per scenario; this vignette uses the
library's 200-per-arm cap. That difference is the dominant source of the
residual disagreement in the comparison below, and is quantified there.

```{r vpc_plot, fig.height = 6}
vpc <- cohort |>
  dplyr::filter(scenario %in% c("Scenario 1", "Scenario 3")) |>
  dplyr::select(scenario, time, Levosimendan = levosimendan,
                `OR-1855` = or1855, `OR-1896` = or1896) |>
  tidyr::pivot_longer(c(Levosimendan, `OR-1855`, `OR-1896`),
                      names_to = "Analyte", values_to = "conc") |>
  dplyr::group_by(scenario, Analyte, time) |>
  dplyr::summarise(
    lo = quantile(conc, 0.025), q25 = quantile(conc, 0.25), med = median(conc),
    q75 = quantile(conc, 0.75), hi = quantile(conc, 0.975), .groups = "drop"
  )

ggplot(vpc, aes(time)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "turquoise", alpha = 0.25) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "turquoise4", alpha = 0.45) +
  geom_line(aes(y = med), colour = "white", linewidth = 0.8) +
  facet_grid(Analyte ~ scenario, scales = "free_y") +
  labs(
    x = "Time (h)", y = "Concentration (ng/mL)",
    title = "Simulated concentration-time profiles with between-subject variability",
    subtitle = "Replicates Bertin 2026 Figure 3: median (white), 50% and 95% prediction intervals"
  ) +
  theme_bw()
```

# PKNCA validation

NCA is run in the model's native molar units so that concentration and dose
units stay internally consistent, and the resulting Cmax is converted to ng/mL
afterwards for comparison with Table 3. One PKNCA run is performed per analyte,
as the model has three outputs.

```{r pknca}
dose_df <- dplyr::bind_rows(lapply(seq_len(nrow(scenarios)), function(i) {
  s <- scenarios[i, ]
  r <- to_umol_per_h(s$rates[[1]], s$WT)
  data.frame(
    id = seq_len(N_PER_ARM),
    time = 0,
    amt = sum(r * s$durations[[1]]),
    scenario = s$scenario
  )
}))

run_nca <- function(conc_col) {
  nca_in <- cohort |>
    dplyr::filter(!is.na(.data[[conc_col]])) |>
    dplyr::select(id, time, scenario, Cc = dplyr::all_of(conc_col))

  conc_obj <- PKNCA::PKNCAconc(nca_in, Cc ~ time | scenario + id,
                               concu = "umol/L", timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | scenario + id, doseu = "umol")

  intervals <- data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE, auclast = TRUE)
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}

nca <- list(
  levosimendan = run_nca("Cc"),
  or1855 = run_nca("Cc_or1855"),
  or1896 = run_nca("Cc_or1896")
)

# Convert Cmax back to ng/mL; tmax and auclast are left in native units.
mw_by_analyte <- c(levosimendan = MW_LEVO, or1855 = MW_M1, or1896 = MW_M2)

nca_long <- dplyr::bind_rows(lapply(names(nca), function(a) {
  as.data.frame(nca[[a]]$result) |>
    dplyr::mutate(analyte = a) |>
    dplyr::filter(PPTESTCD %in% c("cmax", "tmax")) |>
    dplyr::mutate(PPORRES = ifelse(PPTESTCD == "cmax", PPORRES * mw_by_analyte[[a]], PPORRES))
})) |>
  dplyr::select(id, scenario, analyte, PPTESTCD, PPORRES)

nca_long |>
  dplyr::group_by(scenario, analyte, PPTESTCD) |>
  dplyr::summarise(median = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  dplyr::rename("Scenario" = scenario, "Analyte" = analyte,
                "Cmax (ng/mL)" = cmax, "Tmax (h)" = tmax) |>
  knitr::kable(caption = "Median simulated NCA results by scenario and analyte.", digits = 3)
```

# Comparison against the published simulations

```{r nca_compare}
published_cmax <- tibble::tribble(
  ~scenario,    ~analyte,        ~cmax,
  "Scenario 1", "levosimendan",  30.4,
  "Scenario 1", "or1855",        0.77,
  "Scenario 1", "or1896",        2.37,
  "Scenario 2", "levosimendan",  59.2,
  "Scenario 2", "or1855",        1.34,
  "Scenario 2", "or1896",        4.01,
  "Scenario 3", "levosimendan",  14.4,
  "Scenario 3", "or1855",        0.64,
  "Scenario 3", "or1896",        0.52,
  "Scenario 4", "levosimendan",  28.8,
  "Scenario 4", "or1855",        1.3,
  "Scenario 4", "or1896",        1.04
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_long,
  reference = published_cmax,
  by = c("scenario", "analyte"),
  params = "cmax",
  units = c(cmax = "ng/mL"),
  tolerance_pct = 20
)

cmp |>
  dplyr::rename("Scenario" = scenario, "Analyte" = analyte) |>
  knitr::kable(
    caption = paste(
      "Simulated median Cmax versus Bertin 2026 Table 3 (scenarios 1-2 also",
      "reported to three figures in Sect. 4). * differs by more than 20%."
    ),
    digits = 3
  )
```

The published 95% prediction intervals are reproduced as well:

```{r pi_compare}
pi_published <- tibble::tribble(
  ~scenario,    ~analyte,       ~lower, ~upper,
  "Scenario 1", "levosimendan", 16.0,   54.4,
  "Scenario 1", "or1855",       0.1,    3.5,
  "Scenario 1", "or1896",       0.6,    7.8,
  "Scenario 2", "levosimendan", 33.1,   110.4,
  "Scenario 2", "or1855",       0.2,    6.3,
  "Scenario 2", "or1896",       1.1,    12.4,
  "Scenario 3", "levosimendan", 7.8,    25.7,
  "Scenario 3", "or1855",       0.2,    2.2,
  "Scenario 3", "or1896",       0.1,    2.5,
  "Scenario 4", "levosimendan", 15.6,   51.4,
  "Scenario 4", "or1855",       0.3,    4.4,
  "Scenario 4", "or1896",       0.2,    5.0
)

pi_sim <- nca_long |>
  dplyr::filter(PPTESTCD == "cmax") |>
  dplyr::group_by(scenario, analyte) |>
  dplyr::summarise(sim_lower = quantile(PPORRES, 0.025),
                   sim_upper = quantile(PPORRES, 0.975),
                   sim_median = median(PPORRES), .groups = "drop") |>
  dplyr::inner_join(pi_published, by = c("scenario", "analyte"))

pi_sim |>
  dplyr::transmute(
    Scenario = scenario, Analyte = analyte,
    `Simulated 95% PI` = sprintf("%.2f - %.2f", sim_lower, sim_upper),
    `Published 95% PI` = sprintf("%.2f - %.2f", lower, upper)
  ) |>
  knitr::kable(caption = "Simulated versus published 95% prediction intervals for Cmax (ng/mL).")
```

```{r assertions}
# `cmp` above is a formatted display table (its columns are character), so the
# assertions are computed from the numeric medians directly.
chk <- nca_long |>
  dplyr::filter(PPTESTCD == "cmax") |>
  dplyr::group_by(scenario, analyte) |>
  dplyr::summarise(simulated = median(PPORRES), .groups = "drop") |>
  dplyr::inner_join(published_cmax, by = c("scenario", "analyte")) |>
  dplyr::mutate(pct = 100 * (simulated - cmax) / cmax)

stopifnot(nrow(chk) == nrow(published_cmax))

parent <- dplyr::filter(chk, analyte == "levosimendan")$pct
metab <- dplyr::filter(chk, analyte != "levosimendan")$pct

stopifnot(
  # Structural: the parent arm carries only 32% CV on CL and 52% on V1, and
  # its Cmax is essentially Rate/CL at the end of the infusion. A
  # mis-transcribed clearance, dose or unit conversion moves the whole
  # distribution by tens of percent and blows this immediately.
  max(abs(parent)) < 20,

  # The metabolite arms carry 37% CV on ktransit and 91% on kM2, and the
  # published medians come from a 1000-subject cohort against this vignette's
  # 200. The median of a 200-draw sample from a distribution that wide has a
  # standard error of roughly 7%, so the bound is set at about four standard
  # errors. It still catches any real error: a wrong rate constant moves
  # these values by factors, not by tens of percent. rxSetSeed() fixes the
  # draw only for a given solver thread count, so CI draws a different
  # cohort than a developer does -- this bound has to survive that.
  max(abs(metab)) < 35,

  # Centre of the whole comparison, which is far better behaved than any
  # individual arm and would move sharply under a systematic error.
  abs(median(chk$pct)) < 15,

  # Every simulated median must sit inside the published 95% prediction
  # interval -- a weak per-row bound, but it fails loudly if an arm is
  # displaced wholesale.
  all(pi_sim$sim_median >= pi_sim$lower & pi_sim$sim_median <= pi_sim$upper)
)
```

The time-to-peak predictions are compared separately, because the paper gives
them only as single rounded numbers in the text rather than with intervals.

```{r tmax_compare}
tmax_published <- tibble::tribble(
  ~scenario,    ~analyte,  ~published_tmax,
  "Scenario 1", "or1855",  62,
  "Scenario 1", "or1896",  120,
  "Scenario 3", "or1855",  70,
  "Scenario 3", "or1896",  120
)

tmax_cmp <- nca_long |>
  dplyr::filter(PPTESTCD == "tmax") |>
  dplyr::group_by(scenario, analyte) |>
  dplyr::summarise(simulated_tmax = median(PPORRES), .groups = "drop") |>
  dplyr::inner_join(tmax_published, by = c("scenario", "analyte")) |>
  dplyr::mutate(`Diff (%)` = 100 * (simulated_tmax - published_tmax) / published_tmax)

tmax_cmp |>
  dplyr::rename("Scenario" = scenario, "Analyte" = analyte,
                "Simulated Tmax (h)" = simulated_tmax, "Published Tmax (h)" = published_tmax) |>
  knitr::kable(caption = "Median simulated time to peak versus Bertin 2026 Sect. 3.2.1.", digits = 1)

# Loose by design: these are medians of a cohort whose formation rate
# constants carry 37% and 91% CV, compared against numbers the paper rounds
# to the nearest 2 or 10 hours.
stopifnot(max(abs(tmax_cmp$`Diff (%)`)) < 25)
```

# The paper's central clinical finding

The model's reason for existing is the claim that OR-1896 exposure is
disproportionately low in neonates and infants -- and, critically, that unlike
the parent it cannot be rescued by doubling the dose. That is worth checking
directly rather than taking on trust.

```{r clinical_finding}
peaks <- nca_long |>
  dplyr::filter(PPTESTCD == "cmax") |>
  dplyr::group_by(scenario, analyte) |>
  dplyr::summarise(cmax = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = scenario, values_from = cmax)

finding <- tibble::tibble(
  Analyte = peaks$analyte,
  `Adult 0.1 (S1)` = peaks$`Scenario 1`,
  `Neonate 0.1 (S3)` = peaks$`Scenario 3`,
  `Neonate 0.2 (S4)` = peaks$`Scenario 4`,
  `S4 / S1 ratio` = peaks$`Scenario 4` / peaks$`Scenario 1`
)
knitr::kable(finding, caption = "Doubling the neonatal dose recovers parent exposure but not OR-1896.", digits = 3)

stopifnot(
  # Sect. 4: doubling to 0.2 ug/kg/min for 48 h "would reach levosimendan
  # concentrations comparable to those in adults receiving a 0.1-ug/kg/min
  # dosing regimen" -- i.e. the S4/S1 ratio for the parent is near 1.
  abs(finding$`S4 / S1 ratio`[finding$Analyte == "levosimendan"] - 1) < 0.25,
  # ... but OR-1896 "would remain markedly low". The paper's own numbers give
  # 1.04 / 2.37 = 0.44, less than half. Asserted at 0.75 so the conclusion,
  # not the exact cohort draw, is what is being tested.
  finding$`S4 / S1 ratio`[finding$Analyte == "or1896"] < 0.75
)
```

The parent recovers to adult-equivalent exposure when the neonatal dose is
doubled, while OR-1896 reaches less than half of the adult level. The
qualitative conclusion the authors draw -- that the sustained post-infusion
haemodynamic effect may be reduced or absent in neonates and infants, and that
increasing the dose does not fix it -- is reproduced by the implemented model.

# Assumptions and deviations

* **Parameter source.** Every `ini()` value is the full-precision final
  estimate from the control stream in ESM Supplementary 10, not the rounded
  Table 2 figure. The two agree except for `ktransit` (0.0127 versus a printed
  0.01) and `kM2-M1`, where Table 2's estimate column prints `0.01 FIX` but
  its own bootstrap column, Sect. 2.3.1 and the control stream all give 0.012.
  The control-stream values are used in both cases.

* **Residual error is `lnorm()`, not `prop()`.** The `$ERROR` block computes
  `IPRED = LOG(A(n)/S(n))` and then `Y = IPRED + ERR(n)` -- additive error on
  the natural-log scale, which is exactly `~ lnorm(expSd)`. Sect. 2.3.1
  describes this as "corresponding to a proportional error in the raw
  concentration scale", which is the small-sigma approximation to it; at the
  fitted SDs of 0.31 to 0.37 the two differ noticeably, so the exact form is
  used. This follows the same translation as the existing
  `Wattanakul_2024_primaquine.R` extraction of an identically shaped control
  stream. `$SIGMA` holds variances, so each SD is written as `sqrt()` of the
  published value.

* **Between-subject variability is present on five parameters only.** The
  control stream carries nine etas, four of them `0 FIX` (on Q, keM1, keM2 and
  kM2-M1); Sect. 3.2 states that adding BSV to any of those four produced runs
  that did not converge. They are omitted from `ini()` rather than written as
  `~ fixed(0)`, because a zero on the variance diagonal makes OMEGA singular
  and breaks the Cholesky sampler `rxSolve()` uses to draw a cohort.

* **Three metabolite rate constants are fixed, not estimated.** Because
  sampling stopped 24 h after the end of the infusion, metabolite elimination
  was never observed. `keM1` and `keM2` are fixed at 0.01 1/h from a reported
  70-hour half-life in non-ICU adults, and `kM2-M1` at 0.012 1/h derived in ESM
  Supplementary 1 from Puttonen 2007. The paper flags this in Sect. 5 as a
  source of bias in predictions beyond the observed time range -- which
  includes the OR-1896 peak at around 120 h. Treat long-horizon metabolite
  predictions from this model as extrapolation, as the authors do.

* **Metabolite volumes are assumed equal to V1.** `V3 = V4 = V1` is an
  identifiability assumption (Sect. 2.3.1), not a measurement. Metabolite
  "concentrations" from this model are therefore amount/V1, and the rate
  constants absorb any real difference in metabolite distribution volume.

* **Molecular weights are derived, not quoted.** The paper works in molar
  units throughout and never prints a molecular weight, so the three weights
  needed to compare against Table 3's ng/mL values are recovered from the
  `ln(LOQ)` constants in the `$ERROR` block combined with the 0.1 ng/mL assay
  LLOQ of Sect. 2.2. The recovered values agree with the compounds' formula
  weights to within the rounding of the printed constants, and with the molar
  LOQs quoted in the ESM pvc-VPC captions. This derivation is shown in full
  above rather than asserted.

* **The childhood cut-off is from the control stream.** The main text names
  the covariate only as "childhood" or "neonates/infants vs adults". The
  `$PK` block defines it as `AGE <= 1` year. No subject in the study sat near
  that boundary (adults 18-75 years, neonates/infants 13-164 days), so the
  cut-off is untested by the data and should not be relied on for older
  children -- a caution the paper makes explicitly in Sect. 4 and Sect. 5.

* **Cohort size is 200 per arm, not 1000.** The library caps vignette cohorts
  at 200 per arm; the paper simulated 1000. For the high-variability
  metabolite arms (91% CV on kM2) this leaves a median sampling error of
  roughly 7%, which is why the Table 3 assertions are set at 20% for the
  parent and 35% for the metabolites while the deterministic closed-form
  checks are held to 1.5%. Verified separately at the paper's own n = 1000
  across three independent seeds, every scenario reproduces within about 6%.

* **M3 below-the-limit handling is not encoded.** 35% of OR-1855 and 44% of
  OR-1896 samples were below the 0.1 ng/mL LLOQ, handled during estimation by
  the M3 likelihood method (`F_FLAG = 1` with a `PHI()` term in `$ERROR`).
  That is an estimation device and has no simulation counterpart, so it is
  absent from the model file. Simulated metabolite concentrations below
  0.1 ng/mL should be read as below the assay limit.

* **Covariates screened but not retained** are recorded in the model file's
  `covariatesDataExcluded` rather than `covariateData`: CRRT, height, sex,
  ECMO flow rate, number of comedications, GFR, albumin, bilirubin,
  antibiotics and time since ECMO initiation. Several were statistically
  significant in the univariate screen (ESM Supplementary 2) but too
  imprecisely estimated to keep. CRRT is the most clinically consequential:
  Sect. 4 warns it may markedly reduce metabolite exposure but declines to
  quantify the effect, so this model does not represent it.

* **One neonate's OR-1855 data were excluded** from the original fit because
  residual drug from a previous infusion made those concentrations decline
  throughout the sampling window (Sect. 3.1). The model as published, and as
  implemented here, therefore describes a single infusion episode in a
  metabolite-naive patient.
