---
title: "Propofol (Bertin 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Propofol (Bertin 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Bertin_2026_propofol"))$reference`
* Description: `r rxode2::rxode(readModelDb("Bertin_2026_propofol"))$description`
* Article: <https://doi.org/10.1007/s40262-025-01585-2>

# Population

Bertin and colleagues ran a prospective, bicentric, observational PK study in
the adult intensive care units of Lausanne (CHUV) and Geneva (HUG) between
January and December 2023. Forty critically ill adults receiving a continuous
propofol infusion were enrolled: 20 supported by extracorporeal membrane
oxygenation (16 veno-arterial, 3 veno-venous, 1 veno-arteriovenous) and 20
non-ECMO controls, each matched to an ECMO patient on sex, age band, body-weight
band, renal replacement therapy, cirrhosis, total-bilirubin band and cardiac
function (Table 1). The cohort was 87.5% male, median age 57 years (ECMO) and
56 years (controls) over an 18-75 year range, and median body weight 78 kg
(ECMO) and 88 kg (controls) over a 51-120 kg range.

Up to eight samples per patient were drawn over 9 h, giving 300 plasma
concentrations. The final popPK analysis used **289 concentrations from 39
patients**: one control patient's eight concentrations were removed for
unexplained bias, two early concentrations from one ECMO patient were removed on
CWRES grounds, and one sample collected after propofol discontinuation was
removed (Sect. 3.1). Median infusion rates over the sampling period were
2.2 mg/kg/h (ECMO) and 3.0 mg/kg/h (controls), a non-significant difference
(p = 0.14). Sampling began a median of 61 h after the first recorded propofol
dose, so the patients were at or near steady state.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Bertin_2026_propofol")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Bertin_2026_propofol.R`. The table below
collects them in one place for review. Every value comes from the main article;
the supplement was not required.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL at 70 kg) | 68 L/h | Table 2, row `CL (L/h)`; RSE 6.5%, bootstrap median 67, 95% CI 58-76 |
| `lvc` (V1) | 82 L | Table 2, row `V 1 (L)`; RSE 33%, bootstrap median 83, 95% CI 28-169 |
| `lq` (Q) | 26 L/h | Table 2, row `Q (L/h)`; RSE 16%, bootstrap median 26, 95% CI 5-124 |
| `lvp` (V2) | 100 L | Table 2, row `V 2 (L)`; RSE 27%, bootstrap median 101, 95% CI 14-377 |
| `e_wt_cl` (theta BW) | 0.66 | Table 2, row `theta BW`; RSE 38%, bootstrap median 0.69, 95% CI 0.22-1.40 |
| `etalcl` | 0.109392 | Table 2, row `omega CL` = 34 CV%; footnote b gives CV% = sqrt(exp(omega^2) - 1), so omega^2 = log(0.34^2 + 1) |
| `etalvc` | 1.838961 | Table 2, row `omega V 1 (CV%)` = 230; same footnote b conversion, omega^2 = log(2.30^2 + 1) |
| `propSd` | 0.13 | Table 2, row `sigma prop (%)` = 13; RSE 8.8%, bootstrap median 13, 95% CI 11-15 |
| Covariate form `CL_i = CL * (1 + (BW - 70)/70 * theta_BW) * exp(eta_i)` | n/a | Equation printed directly beneath Table 2, with the 70 kg reference restated in the Table 2 legend |
| Linear covariate equation `Param_cov = Param * (1 + (COV - COVmed)/COVmed * theta)` | n/a | Sect. 2.4.1, first displayed equation |
| Two-compartment structure, first-order elimination, parameterised in CL / V1 / Q / V2 | n/a | Sect. 3.1 (preferred over 1-cmt, dOFV = -13.6, p < 0.01; no gain from 3-cmt, dOFV = 0) |
| Proportional residual error | n/a | Sect. 2.4.1 (proportional, additive and mixed compared); Sect. 3.1 |
| No BSV on Q or V2 | n/a | Sect. 3.1: Q dOFV = 0, p > 0.05; V2 dOFV = -3.93, p = 0.05 with unsuccessful RSE, not retained |

# Structural checks at typical values

These checks are deterministic: they compare the packaged model's typical-value
solve (`zeroRe()`) against quantities the paper derives from the same
parameters. Because both sides use identical parameters, the only error is
numerical, so tight bounds are the correct assertion here (unlike the
cohort-derived quantities later in this vignette).

```{r structural-constants}
mod <- readModelDb("Bertin_2026_propofol")
ui <- rxode2::rxode(mod)

# Published point estimates, Table 2.
CL <- 68; V1 <- 82; Q <- 26; V2 <- 100; THETA_BW <- 0.66; REF_WT <- 70

# Micro-constants and the biexponential rate constants of a two-compartment
# mammillary system.
k10 <- CL / V1
k12 <- Q / V1
k21 <- Q / V2
s <- k10 + k12 + k21
p <- k10 * k21
alpha <- (s + sqrt(s^2 - 4 * p)) / 2
beta <- (s - sqrt(s^2 - 4 * p)) / 2

structural <- tibble::tibble(
  Quantity = c(
    "Distribution half-life t1/2 alpha (h)",
    "Elimination half-life t1/2 beta (h)",
    "Steady-state volume Vss = V1 + V2 (L)",
    "Vss per kg at the 70 kg reference (L/kg)",
    "V1 per kg at the 70 kg reference (L/kg)",
    "V2 per kg at the 70 kg reference (L/kg)",
    "CL at 100 kg (L/h)",
    "CL at 70 kg relative to 100 kg (%)"
  ),
  Computed = c(
    log(2) / alpha,
    log(2) / beta,
    V1 + V2,
    (V1 + V2) / REF_WT,
    V1 / REF_WT,
    V2 / REF_WT,
    CL * (1 + (100 - REF_WT) / REF_WT * THETA_BW),
    100 * (1 - CL / (CL * (1 + (100 - REF_WT) / REF_WT * THETA_BW)))
  ),
  Published = c(0.57, 3.90, 182, 2.6, 1.2, 1.4, 87, 22),
  # Each row is pinned to the precision the paper actually printed, rather than
  # to one blanket bound. All of these are deterministic consequences of the
  # Table 2 estimates, so the only slack needed is the published rounding:
  #  * 1% -- Vss (182 L, 3 significant figures) and CL at 100 kg (87 L/h, where
  #    half a unit in the last place is 0.57%).
  #  * 3% -- the two half-lives and the 22% figure. The authors derived the
  #    half-lives from unrounded estimates, while this vignette recomputes them
  #    from the printed ones; Q is given to 2 significant figures, so its
  #    +/- 0.5 L/h (1.9%) propagates straight into beta.
  #  * 5% -- the three L/kg figures, printed to 2 significant figures, where
  #    half a unit in the last place of 1.2 is already 4.2%.
  # A mis-transcribed volume, clearance or unit moves every one of these by tens
  # of percent, so none of these bounds is wide enough to hide one.
  `Tolerance (%)` = c(3, 3, 1, 5, 5, 5, 1, 3),
  `Source` = c(
    "Sect. 3.1", "Sect. 3.1", "Sect. 4", "Sect. 4", "Sect. 4", "Sect. 4",
    "Sect. 3.1", "Sect. 3.1"
  )
) |>
  dplyr::mutate(
    `Difference (%)` = 100 * (Computed - Published) / Published,
    Pass = abs(`Difference (%)`) < `Tolerance (%)`
  )

structural |>
  dplyr::mutate(Computed = round(Computed, 3), `Difference (%)` = round(`Difference (%)`, 2)) |>
  dplyr::relocate(Pass, .after = dplyr::last_col()) |>
  knitr::kable(
    caption = paste(
      "Quantities the paper derives from its own Table 2 estimates,",
      "recomputed from the packaged values. The tolerance on each row is set by",
      "the number of significant figures the paper printed."
    )
  )
```

```{r structural-assert}
stopifnot(nrow(structural) == 8L, all(structural$Pass))
```

```{r closed-form-gate}
# The packaged ODE solve must reproduce the analytic two-compartment
# constant-rate infusion solution. Both sides use the same typical-value
# parameters, so the residual is pure solver error: a tight bound is correct.
rate_mgh <- 2.2 * 70 # median ECMO infusion rate (mg/kg/h) at the 70 kg reference
tinf <- 96

# Materialise the rxEt object with as.data.frame() before adding the covariate
# column: assigning onto an rxEt is silently dropped by rxode2.
ev_typ <- rxode2::et(amt = rate_mgh * tinf, dur = tinf, cmt = "central") |>
  rxode2::et(seq(0, 120, by = 0.25), cmt = "central") |>
  as.data.frame() |>
  dplyr::mutate(WT = REF_WT)

sim_typ <- rxode2::rxSolve(rxode2::zeroRe(mod), events = ev_typ) |>
  as.data.frame()

# Analytic solution: sum of two exponentials driven by a constant-rate infusion
# of duration tinf, with coefficients from the standard two-compartment
# infusion solution (Gibaldi & Perrier).
A <- (alpha - k21) / (V1 * (alpha - beta))
B <- (k21 - beta) / (V1 * (alpha - beta))
conc_analytic <- function(t) {
  during <- pmin(t, tinf)
  infused <- rate_mgh * (
    A / alpha * (1 - exp(-alpha * during)) * exp(-alpha * pmax(t - tinf, 0)) +
      B / beta * (1 - exp(-beta * during)) * exp(-beta * pmax(t - tinf, 0))
  )
  infused
}

cf <- sim_typ |>
  dplyr::filter(time > 0) |>
  dplyr::mutate(
    analytic = conc_analytic(time),
    pct_diff = 100 * (Cc - analytic) / analytic
  )

stopifnot(
  nrow(cf) > 400,
  all(is.finite(cf$pct_diff)),
  max(abs(cf$pct_diff)) < 0.1
)

cat(sprintf(
  "Closed-form gate: %d points, max |difference| = %.4f%% (bound 0.1%%)\n",
  nrow(cf), max(abs(cf$pct_diff))
))
```

The closed-form agreement also rules out the silent failure mode in which
rxode2 collapses a `cl`/`vc` parameterisation onto a one-compartment analytic
solve and discards the peripheral ODE: a one-compartment solve of these
parameters would have a single half-life of `r round(log(2) * 82 / 68, 2)` h
rather than the biexponential `r round(log(2) / alpha, 2)` h /
`r round(log(2) / beta, 2)` h profile matched above.

```{r steady-state-mass-balance}
# At steady state under a constant-rate infusion, Css = Rate / CL. This is a
# mass-balance identity, so it too belongs on the typical-value solve.
css_model <- sim_typ$Cc[sim_typ$time == tinf]
css_expected <- rate_mgh / CL

stopifnot(
  length(css_model) == 1L,
  abs(100 * (css_model - css_expected) / css_expected) < 0.5
)

cat(sprintf(
  "Steady-state Css: model %.4f mg/L vs Rate/CL %.4f mg/L (%.3f%% difference)\n",
  css_model, css_expected, 100 * (css_model - css_expected) / css_expected
))
```

# Virtual cohort

Original observed data are not publicly available. The cohort below mirrors the
Table 1 demographics: two arms of 200 subjects each, an ECMO arm with a median
body weight of 78 kg and a control arm with a median of 88 kg, both truncated to
the observed 51-120 kg range, and each arm given its published median infusion
rate (2.2 and 3.0 mg/kg/h respectively).

The model carries no ECMO term, so the two arms differ only through body weight
and infusion rate. That is exactly the paper's finding, and the figure below is
the check on it.

```{r cohort}
# set.seed() seeds R's RNG, which is what draws the covariates below. It does
# NOT seed rxode2's simulation RNG, and rxode2's streams are partitioned per
# solver thread -- so the eta draws differ between a 2-core CI runner and a
# 16-thread workstation and no seed makes them agree. Every assertion below on a
# cohort-derived quantity is therefore written as a magnitude or a robust
# quantile, never a sign, an exact value, or a bound taken from one run.
set.seed(20260911)

N_PER_ARM <- 200L
TINF <- 96 # h of continuous infusion before the washout window
TEND <- 120 # h; 24 h of washout, about six elimination half-lives

make_arm <- function(n, wt_median, rate_mgkgh, arm, id_offset = 0L) {
  # Log-normal body weight matched to the arm median and truncated to the
  # Table 1 range of 51-120 kg.
  wt <- pmin(pmax(wt_median * exp(stats::rnorm(n, 0, 0.20)), 51), 120)

  subj <- tibble::tibble(
    id = id_offset + seq_len(n),
    WT = wt,
    arm = arm,
    rate_mgkgh = rate_mgkgh
  )

  doses <- subj |>
    dplyr::mutate(
      time = 0,
      # A single constant-rate infusion of TINF hours, expressed as an amount
      # plus a duration.
      amt = rate_mgkgh * WT * TINF,
      dur = TINF,
      evid = 1L,
      cmt = "central"
    )

  obs <- subj |>
    tidyr::crossing(
      time = sort(unique(c(
        seq(0, TINF, by = 2), # steady-state approach and plateau
        seq(TINF, TEND, by = 0.25) # washout, sampled finely for the terminal slope
      )))
    ) |>
    dplyr::mutate(
      amt = NA_real_,
      dur = NA_real_,
      evid = 0L,
      # The ODE state name, never the algebraic observable name "Cc".
      cmt = "central"
    )

  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  make_arm(N_PER_ARM, 78, 2.2, "ECMO", id_offset = 0L),
  make_arm(N_PER_ARM, 88, 3.0, "Control", id_offset = N_PER_ARM)
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep = c("arm", "WT", "rate_mgkgh")
) |>
  as.data.frame()

# rxSolve returns per-subject cl / vc, which later gates compare against.
stopifnot(
  all(c("Cc", "cl", "vc", "arm") %in% names(sim)),
  nrow(sim) > 0,
  !anyNA(sim$Cc)
)
```

```{r covariate-recovery}
# The body-weight effect on CL is a deterministic function of the drawn WT and
# eta, so the per-subject CL that rxSolve reports must reproduce the printed
# equation exactly. This catches a mis-signed or mis-normalised covariate term.
cl_chk <- sim |>
  dplyr::distinct(id, arm, WT, cl) |>
  dplyr::mutate(cl_wt_factor = cl / (1 + (WT - REF_WT) / REF_WT * THETA_BW))

# After dividing out the covariate factor, the remainder is CL * exp(eta), whose
# geometric mean across a log-normal eta converges to the typical value 68 L/h.
gm_cl <- exp(mean(log(cl_chk$cl_wt_factor)))

stopifnot(
  nrow(cl_chk) == 2L * N_PER_ARM,
  all(cl_chk$cl > 0),
  # Geometric mean of 400 draws from a 34% CV log-normal: the standard error of
  # the log mean is 0.331/sqrt(400) = 0.0165, so a 10% band is about 6 SE.
  abs(100 * (gm_cl - CL) / CL) < 10
)

cat(sprintf(
  "Covariate-adjusted geometric-mean CL across the cohort: %.2f L/h (Table 2: %g L/h)\n",
  gm_cl, CL
))
```

# Replicate published figures

```{r figure-1, fig.width = 7, fig.height = 4}
# Replicates Figure 1 of Bertin 2026: observed propofol concentrations corrected
# for the infusion rate and body weight, ECMO (left) vs control (right). In the
# paper the two panels overlap, which is the visual form of the study's central
# negative finding. Because the packaged model contains no ECMO term, any
# residual separation here can only come from the weight difference between the
# arms.
sim |>
  dplyr::filter(time > 0, time <= TINF) |>
  dplyr::mutate(Cc_norm = Cc / rate_mgkgh) |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc_norm, 0.05),
    Q50 = quantile(Cc_norm, 0.50),
    Q95 = quantile(Cc_norm, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~arm) +
  scale_y_log10() +
  labs(
    x = "Time (h)",
    y = "Cc / infusion rate ((mg/L) / (mg/kg/h))",
    title = "Figure 1 - infusion-rate-corrected concentrations by arm",
    caption = "Replicates Figure 1 of Bertin 2026. Median with 5th-95th percentile band."
  )
```

```{r figure-1-assert}
# The paper's finding is that the arms do not differ. Assert the MAGNITUDE of
# the separation, not its sign: with no ECMO term in the model the expected
# separation is the pure weight effect, and the sign of any residual is a coin
# flip that would flicker between thread counts.
arm_med <- sim |>
  dplyr::filter(time >= TINF - 24, time <= TINF) |>
  dplyr::mutate(Cc_norm = Cc / rate_mgkgh) |>
  dplyr::group_by(arm) |>
  dplyr::summarise(med = median(Cc_norm), .groups = "drop")

sep_pct <- 100 * abs(diff(arm_med$med)) / mean(arm_med$med)

# Expected separation from weight alone: an 88 kg subject clears
# (1 + 18/70*0.66) / (1 + 8/70*0.66) = 1.10 times as fast as a 78 kg subject, and
# the rate-normalised steady-state concentration is WT/CL, so the arms should sit
# roughly 10% apart. This is a coarse envelope check -- it confirms the arms
# overlap as Figure 1 shows, but it is NOT what guarantees the absence of an ECMO
# effect, because a 25%-sized effect on top of the weight difference would still
# fall inside it. The structural assertion in the next chunk is what carries that
# guarantee.
stopifnot(sep_pct < 35)

cat(sprintf(
  "Rate-normalised steady-state median separation between arms: %.1f%% (bound 35%%)\n",
  sep_pct
))
```

```{r no-ecmo-term}
# Exact structural check on the paper's central negative finding: the published
# final model carries no ECMO term at all, so neither the parameter table nor the
# model body may mention one. This cannot flicker with the cohort draw, and it is
# the assertion that would go red if a screened-and-rejected covariate were ever
# reinstated in the model file.
# ui$lstExpr is the model() body alone. The description and covariatesDataExcluded
# metadata deliberately DO mention ECMO, so the check must not be run against the
# whole model function.
model_text <- paste(
  vapply(ui$lstExpr, function(e) paste(deparse(e), collapse = " "), character(1)),
  collapse = " ; "
)

stopifnot(
  length(ui$lstExpr) > 0L,
  # The excluded covariates are documented in metadata but must not be used.
  !any(grepl("ECMO", ui$iniDf$name, ignore.case = TRUE)),
  !grepl("ECMO", model_text, ignore.case = TRUE),
  # WT is the one covariate the final model does use.
  grepl("WT", model_text, fixed = TRUE)
)

cat("Structural check: no ECMO term in ini() or model(); WT is the sole covariate.\n")
```

```{r figure-3, fig.width = 7, fig.height = 4}
# Analogue of Figure 3 of Bertin 2026 (prediction-corrected VPC): the model's own
# 5th / 50th / 95th prediction percentiles, which the published observed
# percentiles fall within. Plotted here on the simulated scale, including the
# washout phase the study design did not sample.
sim |>
  dplyr::filter(time > 0) |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  geom_vline(xintercept = TINF, linetype = "dashed") +
  facet_wrap(~arm) +
  scale_y_log10() +
  labs(
    x = "Time (h)",
    y = "Propofol plasma concentration (mg/L)",
    title = "Prediction interval, continuous infusion then washout",
    caption = paste(
      "Analogue of Figure 3 of Bertin 2026. Median with 5th-95th percentile band;",
      "dashed line marks the end of the 96 h infusion."
    )
  )
```

# PKNCA validation

The primary check is the exact dose-clearance identity `AUC_0-inf = Dose / CL`,
which holds for every subject regardless of compartment structure, volume, or
whether steady state was reached. A steady-state check over the last 24 h of
infusion is run alongside it, restricted to the subjects for whom steady state
is physically attainable within the simulated infusion. Finally the terminal
half-life is recovered from the typical-value washout and compared against the
value Sect. 3.1 publishes.

```{r pknca-setup}
# Only !is.na(Cc) -- adding time > 0 or Cc > 0 would drop the interval-anchor
# rows PKNCA needs.
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

# Guarantee a time = 0 record per subject; propofol is given intravenously and
# the cohort starts drug-free, so Cc = 0 there.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, arm, time)

stopifnot(all(sim_nca$Cc >= 0), nrow(sim_nca) > 0)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

# The infusion is one event of TINF hours; for the steady-state interval the
# relevant dose is the amount delivered during that interval.
dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, arm)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)
```

```{r subject-half-lives}
# Per-subject terminal half-life, derived from the model's own reported cl and
# vc plus the two parameters that carry no IIV (Q and V2 are fixed across
# subjects). Used below to decide which subjects can have reached steady state.
per_subject <- sim |>
  dplyr::distinct(id, arm, WT, rate_mgkgh, cl, vc) |>
  dplyr::mutate(
    k10_i = cl / vc,
    k12_i = Q / vc,
    k21_i = Q / V2,
    beta_i = ((k10_i + k12_i + k21_i) -
      sqrt((k10_i + k12_i + k21_i)^2 - 4 * k10_i * k21_i)) / 2,
    thalf_i = log(2) / beta_i
  )

stopifnot(nrow(per_subject) == 2L * N_PER_ARM, all(is.finite(per_subject$thalf_i)))
```

```{r pknca-dose-identity}
# PRIMARY GATE. AUC_0-inf = Dose / CL is an exact identity for an intravenous
# dose, independent of the number of compartments, of the volumes, and of
# whether steady state was reached. Both sides use the same drawn per-subject
# parameters, so the only residual is trapezoidal-integration and terminal-
# extrapolation error: a tight bound is the correct assertion here.
nca_inf <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(
    conc_obj, dose_obj,
    intervals = data.frame(start = 0, end = Inf, aucinf.obs = TRUE)
  ),
  verbose = FALSE
)

mb <- as.data.frame(nca_inf) |>
  dplyr::select(arm, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::left_join(per_subject, by = c("id", "arm")) |>
  dplyr::mutate(
    cl_nca = rate_mgkgh * WT * TINF / aucinf.obs,
    pct_diff = 100 * (cl_nca - cl) / cl
  )

stopifnot(
  nrow(mb) == 2L * N_PER_ARM,
  all(is.finite(mb$pct_diff)),
  # Realised max 0.85% over this 400-subject draw; the error is numerical, not
  # cohort-driven, but 2% leaves headroom for a draw that puts more subjects in
  # the long-half-life tail where the terminal extrapolation does more work.
  max(abs(mb$pct_diff)) < 2
)

cat(sprintf(
  "Dose/AUCinf identity: max |CL(NCA) - CL(model)| = %.3f%% over %d subjects (bound 2%%)\n",
  max(abs(mb$pct_diff)), nrow(mb)
))
```

```{r pknca-steady-state}
intervals_ss <- data.frame(
  start = TINF - 24,
  end = TINF,
  auclast = TRUE,
  cmax = TRUE,
  cmin = TRUE
)

nca_ss <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals_ss),
  verbose = FALSE
)

ss <- as.data.frame(nca_ss) |>
  dplyr::select(arm, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

stopifnot(nrow(ss) == 2L * N_PER_ARM, !anyNA(ss$auclast))

# SECONDARY GATE. At steady state the amount infused over an interval equals
# the amount cleared, so CL = (rate * 24) / AUCtau. Unlike the identity above,
# this one requires the subject to actually BE at steady state, and the
# published 230% CV on V1 means a minority are not: a subject drawing
# V1 = 3,000 L has a terminal half-life near 60 h and is still accumulating at
# the end of a 96 h infusion.
#
# The subset condition is a stated physical criterion, not a fit to the data:
# eight terminal half-lives of infusion puts a subject within 0.4% of steady
# state, so t1/2 < TINF/8 = 12 h qualifies. The size of the qualifying subset is
# asserted as well, so this gate cannot pass vacuously on an empty filter.
ss_mb <- ss |>
  dplyr::left_join(per_subject, by = c("id", "arm")) |>
  dplyr::mutate(
    at_steady_state = thalf_i < TINF / 8,
    cl_nca = rate_mgkgh * WT * 24 / auclast,
    pct_diff = 100 * (cl_nca - cl) / cl
  )

atss <- ss_mb[ss_mb$at_steady_state, ]

stopifnot(
  # Guard against a vacuous pass: the criterion must select most of the cohort.
  nrow(atss) >= 0.80 * nrow(ss_mb),
  all(is.finite(atss$pct_diff)),
  max(abs(atss$pct_diff)) < 1
)

cat(sprintf(
  paste0(
    "Steady-state mass balance: max |CL(NCA) - CL(model)| = %.4f%% over the %d of %d\n",
    "subjects with a terminal half-life under %g h (bound 1%%).\n",
    "The %d excluded subjects have half-lives of %.0f-%.0f h and are still\n",
    "accumulating at 96 h -- a direct consequence of the published 230%% CV on V1.\n"
  ),
  max(abs(atss$pct_diff)), nrow(atss), nrow(ss_mb), TINF / 8,
  sum(!ss_mb$at_steady_state),
  min(ss_mb$thalf_i[!ss_mb$at_steady_state]),
  max(ss_mb$thalf_i[!ss_mb$at_steady_state])
))
```

```{r pknca-halflife}
# The published half-lives are typical-value quantities derived from Table 2, so
# the gate on them belongs on the zeroRe() solve, not on a cohort whose V1 has a
# 230% CV. The cohort half-life is reported alongside for context only.
conc_typ <- sim_typ |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::transmute(id = 1L, time, Cc, arm = "Typical (70 kg)")

dose_typ <- tibble::tibble(
  id = 1L, time = 0, amt = rate_mgh * tinf, arm = "Typical (70 kg)"
)

nca_typ <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_typ, Cc ~ time | arm + id),
  PKNCA::PKNCAdose(dose_typ, amt ~ time | arm + id),
  # Start after the infusion ends so the fit sees only the washout, and start
  # 4 h in so the alpha phase (t1/2 = 0.56 h, i.e. seven half-lives by 4 h) has
  # decayed away and the slope is the true terminal beta phase. Fitting from the
  # moment of withdrawal would include the distribution transient and read long.
  intervals = data.frame(start = tinf + 4, end = TEND, half.life = TRUE)
))

typ_hl <- as.data.frame(nca_typ) |>
  dplyr::filter(PPTESTCD == "half.life") |>
  dplyr::pull(PPORRES)

stopifnot(length(typ_hl) == 1L, is.finite(typ_hl))
```

## Comparison against published values

Bertin 2026 reports no NCA table; the only exposure-derived quantities it
publishes are the two half-lives derived from the final model (Sect. 3.1). The
terminal half-life recovered by PKNCA from the typical-value washout is compared
against the published value below.

```{r nca-compare}
published <- tibble::tribble(
  ~arm, ~half.life,
  "Typical (70 kg)", 3.90
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_typ,
  reference = published,
  by = "arm",
  params = "half.life",
  units = c(half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Terminal half-life recovered by PKNCA from the typical-value washout vs the",
    "value Bertin 2026 derives from the same parameters (Sect. 3.1).",
    "* marks a >20% difference."
  )
)
```

```{r nca-compare-assert}
# Deterministic: both sides derive from the same typical-value parameters, so
# the only error is the published rounding of CL/V1/Q/V2 plus the terminal-slope
# fit. 5% holds that and still fails on a mis-transcribed volume or clearance.
stopifnot(abs(100 * (typ_hl - 3.90) / 3.90) < 5)

cat(sprintf(
  "Typical-value terminal half-life: PKNCA %.3f h vs published %.2f h (%.2f%% difference)\n",
  typ_hl, 3.90, 100 * (typ_hl - 3.90) / 3.90
))
```

```{r cohort-context}
# Cohort context only -- not a gate. With a 230% CV on V1 the per-subject
# half-life distribution is extremely wide, which is the paper's point about
# propofol variability in this population.
ss |>
  dplyr::left_join(sim |> dplyr::distinct(id, WT, rate_mgkgh), by = "id") |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    `Median Css (mg/L)` = median(cmax),
    `5th percentile Css (mg/L)` = quantile(cmax, 0.05),
    `95th percentile Css (mg/L)` = quantile(cmax, 0.95),
    .groups = "drop"
  ) |>
  dplyr::rename("Arm" = arm) |>
  knitr::kable(
    digits = 2,
    caption = paste(
      "Simulated steady-state concentrations by arm over the last 24 h of",
      "infusion. The width of the interval reflects the published 34% CV on CL",
      "and 230% CV on V1."
    )
  )
```

# Assumptions and deviations

- **The 70 kg normalising constant is the paper's, not an assumption.** The
  Methods define the linear covariate form with `COVmed` as "the corresponding
  median value in the study population", but the Table 2 equation and legend
  both instantiate it at 70 kg, which is *not* the cohort median (78 kg on ECMO,
  88 kg in controls). Three independent cross-checks confirm 70 kg is what was
  used: the paper's own worked example gives 87 L/h at 100 kg (the model gives
  87.2; a median-normalised form would give 86.2), and Sect. 4 quotes V1 = 82 L
  as 1.2 L/kg and V2 = 100 L as 1.4 L/kg, both of which are the L values divided
  by 70.
- **ECMO is not in the model.** It was the study's primary covariate of
  interest, was significant on CL and V1 in initial screening, and was dropped
  after a sensitivity analysis showed the effect was driven by two leverage
  points among the controls (Sect. 3.1). It is recorded in the model file's
  `covariatesDataExcluded` list rather than `covariateData`, along with the six
  other screened-and-rejected covariates (time since ECMO initiation, albumin,
  total bilirubin, height, age, sex, cirrhosis). Simulating an ECMO effect from
  this model is therefore not possible, which is the correct behaviour.
- **No BSV on Q or V2.** The paper tested both and retained neither
  (Sect. 3.1), so their absence is a published modelling decision rather than an
  unreported variance. They are consequently omitted rather than encoded as
  `fixed(0)`.
- **`n_subjects` is 39, not 40.** Forty patients were enrolled and are described
  in Table 1, but the final popPK model was fit to 289 concentrations from 39
  patients after the documented exclusions (Sect. 3.1). The `population`
  metadata records the analysis population and notes the distinction.
- **Virtual-cohort distributions are assumptions.** The paper publishes only
  medians and ranges for body weight, so the cohort above uses a log-normal
  weight distribution matched to each arm's median and truncated to the observed
  51-120 kg range. Infusion rates are held at each arm's published median rather
  than sampled, and the boluses that 45-70% of patients received are not
  reproduced. None of these feed a gate: every assertion is either a
  typical-value quantity or a per-subject identity that holds for any covariate
  draw.
- **The simulated regimen is longer than the study's.** Patients were sampled
  over 9 h, a median of 61 h into their infusion. The cohort here runs a 96 h
  infusion followed by a 24 h washout so that the steady-state and
  terminal-phase gates both have a window to measure in. The study design
  contains no washout at all.
- **No non-paper-derived parameter values.** Every value in the model file comes
  from Table 2 or the Methods of the main article. The supplement (which holds
  the final NONMEM control stream as Supplementary Table 1, plus the detailed
  covariate-analysis tables) was not needed and was not retrieved: the final
  model equation is printed in full beneath Table 2 and every estimate is in
  Table 2 itself.
