---
title: "Depot testosterone cypionate and the HPG axis (Bi 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Depot testosterone cypionate and the HPG axis (Bi 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5
)
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(tidyr)
library(ggplot2)
rxode2::rxSetSeed(20180301)
```

# The paper

Bi Y, Perry PJ, Ellerby M, Murry DJ. *Population Pharmacokinetic/Pharmacodynamic
Modeling of Depot Testosterone Cypionate in Healthy Male Subjects.*
CPT Pharmacometrics Syst Pharmacol 2018;7(4):259-268.
[doi:10.1002/psp4.12287](https://doi.org/10.1002/psp4.12287)

Thirty-one healthy men were randomised to 100, 250 or 500 mg/week of
intramuscular testosterone cypionate (TC) and received 14 consecutive weekly
injections (study weeks 2-15), bracketed by two weeks of placebo injections
before and twelve weeks of placebo injections after. The 250 and 500 mg arms
deliberately mimic supratherapeutic doses used illicitly rather than
replacement-therapy doses. The paper builds a sequential PK/PD model linking
testosterone exposure to suppression of luteinizing hormone (LH) and of
spermatogenesis, and to the recovery of both after dosing stops.

The analysis is unusual in that the drug and the endogenous hormone are the
**same molecule**: exogenous TC adds to an endogenous testosterone pool whose
own secretion rate is regulated by LH through the hypothalamic-pituitary-gonadal
(HPG) feedback loop, and which the exogenous drug suppresses. The paper states
plainly why this defeats conventional analysis: "It is difficult to calculate the
PK parameters of TC using traditional noncompartmental methods, especially when
the endogenous testosterone secretion rate is suppressed during the course of TC
administration."

## What this article contributes to nlmixr2lib

```{r models}
mods <- c("Bi_2018_testosteroneCypionate", "Bi_2018_testosteroneCypionate_sperm")
uis <- lapply(mods, function(n) rxode2::rxode(nlmixr2lib::readModelDb(n)))
names(uis) <- mods

tibble::tibble(
  Model = mods,
  Endpoints = vapply(uis, function(u) paste(u$predDf$var, collapse = ", "), character(1)),
  States = vapply(uis, function(u) paste(u$props$cmt, collapse = ", "), character(1))
) |>
  knitr::kable(caption = "The two nlmixr2lib entries contributed by Bi 2018.")
```

`Bi_2018_testosteroneCypionate` couples the two model layers the paper fitted
sequentially. Bi 2018 fitted the testosterone PK with the **post hoc LH
prediction supplied as a data column** ("The individual post hoc LH prediction
from the PD model was used in this indirect model because of different sampling
time of testosterone and LH"), and fitted the LH model with the post hoc PK
parameters fixed. That is an estimation strategy for two endpoints sampled on
different schedules, not two independent models: the supplementary LH control
stream already carries the absorption, testosterone, effect-site and LH
differential equations in a single `$DES` block. Encoding them as one model
closes the feedback loop that the sequential fit approximates, so the library
entry is simulatable without an externally supplied LH time course.

`Bi_2018_testosteroneCypionate_sperm` is kept separate, exactly as the authors
built it: it is driven not by a state of the PK model but by `CAV`, a rolling
18-week average testosterone concentration that Bi 2018 pre-computed and passed
in as a regressor. That structure is reproduced faithfully here, and this
vignette shows how to generate `CAV` from the PK model.

The sperm-motility fit (Supplementary Table S1) and the LH60 / LH120 fits are
described by the paper as sensitivity analyses with "similar parameter
estimation and comparable performance", so per nlmixr2lib policy they are not
extracted.

# Population

```{r population}
pop <- uis[[1]]$population
tibble::tibble(Field = names(pop), Value = vapply(pop, function(x) {
  paste(format(x), collapse = "; ")
}, character(1))) |>
  knitr::kable(caption = "Population metadata (Bi 2018 Table 1 and Methods).")
```

# Source trace

Every structural equation and every parameter value, with where it came from.
Bi 2018 prints its differential equations in the Results text but places the
covariate functional forms, the centring constants and the residual-error
parameterisation only in the Supplementary Material NONMEM control streams
(file `PSP4-7-259-s006.docx`, referred to below as **S6**).

```{r source-trace}
tibble::tribble(
  ~Component, ~Encoding, ~Source,
  "Absorption", "d/dt(depot) = -ka * depot", "Results, first displayed equation; S6 PK $DES",
  "Testosterone", "d/dt(central) = ksec_te + ka*depot - (cl/vc)*central", "Results, first displayed equation; S6 PK $DES",
  "Endogenous secretion", "ksec_te = kin_te + emax_te*LH^hill/(LH^hill + ec50_te^hill)", "Results, third displayed equation; S6 PK $PK (k1)",
  "Effect site", "d/dt(effect) = ke0 * (Cc - effect)", "S6 LH $DES (DADT(3))",
  "LH30", "d/dt(lh30) = kin_lh*(1 - INH) - kout_lh*lh30", "Results, LH section displayed equation; S6 LH $DES",
  "LH30 inhibition", "INH = Ce^hill_lh / (Ce^hill_lh + ic50_lh^hill_lh)", "Results, LH section; Emax fixed to 1 per the text",
  "Sperm", "d/dt(sperm) = kin_sperm*remaining - kout_sperm*sperm", "S6 sperm-count $DES",
  "Sperm inhibition", "remaining = 1 - emax_sperm*CAV^h/(CAV^h + ic50_sperm^h)", "Results, spermatogenesis section; S6 sperm-count $PK",
  "Emax constraint", "emax_sperm = expit(logitemax_sperm)", "Results: 'Logistic transformation was applied on Emax'",
  "CL covariates", "(WT_BASE/85)^0.785 * exp(0.016*(WT - WT_BASE - 2.90))", "Table 2; forms and centring from S6 PK $PK",
  "V covariates", "(WT_BASE/85)^1.71 * (ALB_BASE/4.60)^-1.55 * (1 - 0.27*(dALB + 0.2))", "Table 2; forms and centring from S6 PK $PK",
  "LH30 covariates", "ic50 * (WT_BASE/84.70)^-1.14 ; kin * (T4/7.40)^1.19", "Table 3; references from S6 LH $PK",
  "Sperm covariate", "ic50_sperm * (WT_BASE/85)^-1.27", "Table 3; see Errata on the reference weight",
  "PK residual", "add(0.511) + prop(0.239)", "S6 PK: W = sqrt(IPRED^2*THETA(8) + THETA(9)), so Table 2's sigma^2 rows are variances",
  "PD residuals", "add() on log(state + 1)", "S6 LH / sperm: W = THETA(6) directly, so Table 3's sigma^2 rows are SDs"
) |>
  knitr::kable(caption = "Source trace for the structural model.")
```

```{r source-trace-params}
ini1 <- as.data.frame(uis[[1]]$iniDf)
ini2 <- as.data.frame(uis[[2]]$iniDf)
dplyr::bind_rows(
  ini1 |> dplyr::mutate(Model = "PK + LH"),
  ini2 |> dplyr::mutate(Model = "Sperm")
) |>
  dplyr::filter(!is.na(est)) |>
  dplyr::select(Model, Parameter = name, Estimate = est, Fixed = fix, Label = label) |>
  knitr::kable(digits = 6, caption = "All estimated quantities in both model files.")
```

# Deterministic structural checks

These checks involve no random effects, so they are exact identities of the
encoded model and are asserted tightly.

## Typical-value parameters reproduce Table 2

Bi 2018 reports "CL/F and V/F for tT were estimated to be 2.6 (kL/day) and
14.4 kL in healthy men with **median covariate values**". Because both
change-from-baseline covariates are centred away from zero -- weight at
**+2.90 kg** and albumin at **-0.2 g/dL**, the cohort medians of the *changes*
over 14 weeks of androgen dosing -- a subject whose weight and albumin have not
moved is *not* the typical subject. Reproducing 2.6 and 14.4 requires supplying
the median covariate movement, which is a direct test that the centring
constants were transcribed correctly.

```{r typical-values}
covs_median <- list(WT_BASE = 85, WT = 85 + 2.90, ALB_BASE = 46.0, ALB = 46.0 - 2.0, T4 = 7.40)
covs_nochange <- list(WT_BASE = 85, WT = 85, ALB_BASE = 46.0, ALB = 46.0, T4 = 7.40)

solve_pk <- function(ui, times, doses = NULL, covs, nsub = NULL, ...) {
  ev <- if (is.null(doses)) rxode2::et(times) else {
    rxode2::et(amt = doses$amt, cmt = "depot", time = doses$time,
               ii = doses$ii, addl = doses$addl) |> rxode2::et(times)
  }
  d <- as.data.frame(ev)
  d$dvid <- ifelse(d$evid == 0, 1L, NA_integer_)
  for (nm in names(covs)) d[[nm]] <- covs[[nm]]
  args <- list(object = ui, events = d, returnType = "data.frame", useLinCmt = FALSE, ...)
  if (!is.null(nsub)) args$nSub <- nsub
  do.call(rxode2::rxSolve, args)
}

pk_typ <- rxode2::zeroRe(uis[[1]])
chk_med <- solve_pk(pk_typ, c(0, 1), covs = covs_median)
chk_non <- solve_pk(pk_typ, c(0, 1), covs = covs_nochange)

tv <- tibble::tibble(
  Quantity = c("CL/F (kL/day)", "V/F (kL)"),
  `Bi 2018 Table 2` = c(2.6, 14.4),
  `At median covariate change` = c(chk_med$cl[1], chk_med$vc[1]),
  `At zero covariate change` = c(chk_non$cl[1], chk_non$vc[1])
)
knitr::kable(tv, digits = 4, caption = "Typical CL/F and V/F.")

stopifnot(
  abs(chk_med$cl[1] - 2.6) < 0.005,
  abs(chk_med$vc[1] - 14.4) < 0.02
)
```

## Baseline identities

```{r baselines}
b1 <- solve_pk(pk_typ, c(0, 1), covs = covs_median)
sp_typ <- rxode2::zeroRe(uis[[2]])
sp0 <- rxode2::rxSolve(
  sp_typ,
  data.frame(id = 1L, time = c(0, 1), evid = 0L, amt = NA_real_, dvid = 1L,
             CAV = 1e-9, WT_BASE = 85),
  returnType = "data.frame", useLinCmt = FALSE
)

base_tab <- tibble::tibble(
  Identity = c(
    "Cc(0) = estimated baseline tT (ng/mL)",
    "LH30(0) = kin_lh / kout_lh (IU/L)",
    "effect(0) = Cc(0) (ng/mL)",
    "sperm(0) = kin_sperm / kout_sperm",
    "emax_sperm = expit(4.65)"
  ),
  Expected = c(6.24, 1.7 / 0.11, 6.24, 6.17 / 0.0696, plogis(4.65)),
  Model = c(b1$Cc[1], b1$lh30[1], b1$effect[1], sp0$sperm[1], sp0$emax_sperm[1])
)
knitr::kable(base_tab, digits = 4, caption = "Initial-condition identities.")
stopifnot(max(abs(base_tab$Expected - base_tab$Model)) < 1e-6)
```

The endogenous secretion rate implied at baseline is a quantity Bi 2018 reports
independently in the Discussion: "The mean post hoc endogenous testosterone
secretion rate at baseline in PK models for tT models was 12.5 mg/day."

```{r baseline-secretion}
ksec0 <- b1$ksec_te[1]
cat(sprintf("Model baseline secretion rate: %.2f mg/day (Bi 2018 Discussion: 12.5 mg/day)\n", ksec0))
# The paper's 12.5 is a mean over post hoc individual parameters; the typical-value
# prediction need not equal it. A transcription error in b_base, Emax, LH50 or the
# LH baseline would move this by tens of percent.
stopifnot(abs(ksec0 - 12.5) / 12.5 < 0.15)
```

## Covariate effects reproduce the Figure 1 forest plot

Bi 2018 quotes four fold-changes for a 110 kg subject (the 95th percentile)
relative to the 85 kg typical subject. These are reference-invariant ratios and
so test the exponents directly.

```{r forest}
fold <- function(exponent, ref = 85) (110 / ref)^exponent
forest <- tibble::tibble(
  Effect = c("CL/F", "V/F", "LH30 potency (TC50)", "Sperm potency (Cavg50)"),
  `Bi 2018 quoted` = c(1.23, 1.58, 0.74, 0.72),
  `Source` = c("Results, PK section", "Results, PK section",
               "Results, LH section", "Results, spermatogenesis section"),
  `Model` = c(fold(0.785), fold(1.71), fold(-1.14, 84.70) / (85 / 84.70)^-1.14, fold(-1.27))
) |>
  dplyr::mutate(`% diff` = 100 * (Model - `Bi 2018 quoted`) / `Bi 2018 quoted`)
knitr::kable(forest, digits = 3, caption = "Figure 1 forest-plot fold-changes at 110 kg vs 85 kg.")

# V/F is the loosest: Bi 2018's quoted 1.58 corresponds to the BOOTSTRAP median
# exponent 1.77 ((110/85)^1.77 = 1.578) rather than the final estimate 1.71
# ((110/85)^1.71 = 1.554), a 1.5% gap. All four are deterministic.
stopifnot(all(abs(forest$`% diff`) < 3))
```

# Exogenous testosterone pharmacokinetics

To validate the drug-disposition layer on its own, the endogenous secretion and
the baseline initial condition are switched off by re-specifying three `ini()`
entries. What remains is an ordinary one-compartment first-order-absorption
model, for which mass balance must hold exactly.

```{r exogenous-model}
exo <- uis[[1]] |>
  rxode2::ini(lkin_te = log(1e-12), lemax_te = log(1e-12), lrbase_te = log(1e-12))
exo_typ <- rxode2::zeroRe(exo)

doses <- c(100, 250, 500)
nca_times <- sort(unique(c(seq(0, 5, by = 0.05), seq(5, 120, by = 0.25))))
sim_exo <- dplyr::bind_rows(lapply(seq_along(doses), function(i) {
  s <- solve_pk(exo_typ, nca_times,
                doses = list(amt = doses[i], time = 0, ii = 0, addl = 0),
                covs = covs_median)
  s$id <- i
  s$treatment <- sprintf("%d mg", doses[i])
  s
}))
```

## Mass balance: `CL/F * AUCinf` must return the dose

```{r mass-balance}
recovery <- sim_exo |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(id, treatment) |>
  dplyr::summarise(
    cl = dplyr::first(cl), vc = dplyr::first(vc),
    auc_last = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    c_last = dplyr::last(Cc), t_last = dplyr::last(time),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    kel = cl / vc,
    aucinf = auc_last + c_last / kel,
    dose = doses,
    recovered = cl * aucinf,
    `% recovered` = 100 * recovered / dose,
    `t1/2 (day)` = log(2) / kel
  )

recovery |>
  dplyr::select(Treatment = treatment, `AUCinf (ng/mL*day)` = aucinf,
                `Dose (mg)` = dose, `CL/F * AUCinf (mg)` = recovered,
                `% recovered`, `t1/2 (day)`) |>
  knitr::kable(digits = 3, caption = "Dose recovery for the exogenous-only model.")

# Exact identity of a linear one-compartment model; only numerical integration
# error separates the two sides.
stopifnot(all(abs(recovery$`% recovered` - 100) < 0.05))
# Dose proportionality: AUCinf / dose must be identical across the three arms.
stopifnot(diff(range(recovery$aucinf / recovery$dose)) < 1e-6)
```

The volume is recovered the same way, confirming that the `mg / kL = ng/mL`
unit chain in the model file is self-consistent.

```{r volume-check}
v_implied <- recovery$cl / recovery$kel
cat(sprintf("Implied V/F from CL/F and lambda_z: %.4f kL (model vc: %.4f kL)\n",
            v_implied[1], recovery$vc[1]))
stopifnot(all(abs(v_implied - recovery$vc) < 1e-6))
```

## Noncompartmental analysis with PKNCA

```{r pknca}
sim_nca <- sim_exo |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

dose_df <- tibble::tibble(
  id = seq_along(doses), time = 0, amt = doses,
  treatment = sprintf("%d mg", doses)
)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "ng/mL", timeu = "day")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id, doseu = "mg")

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE, cl.obs = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

Bi 2018 reports no per-arm noncompartmental table -- by design, since the
endogenous pool makes conventional NCA inapplicable to the observed data. The
one NCA-comparable number it prints is the model-based half-life: "Our estimated
post hoc median tT half-life was 4.05 days." The exogenous-only terminal
half-life is compared against it below, with the caveat that the published value
is a median across 31 subjects carrying inter-individual variability on both
CL/F and V/F, whereas the simulated value is the typical-value half-life.

```{r nca-comparison}
published <- tibble::tibble(
  treatment = sprintf("%d mg", doses),
  half.life = 4.05
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "treatment",
  units = c(half.life = "day"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = paste(
  "Simulated exogenous-only terminal half-life vs Bi 2018's post hoc median.",
  "* marks a difference above 20%."
))
```

```{r nca-table}
as.data.frame(nca_res$result) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::rename(
    "Treatment" = treatment, "Cmax (ng/mL)" = cmax, "Tmax (day)" = tmax,
    "AUC0-inf (ng/mL*day)" = aucinf.obs, "t1/2 (day)" = half.life,
    "CL/F (kL/day)" = cl.obs
  ) |>
  knitr::kable(digits = 3, caption = "Full NCA summary for the exogenous-only model.")
```

# The HPG axis under weekly dosing

The full model is now restored and simulated over the trial schedule: 14 weekly
injections from week 2 through week 15, followed by observation to week 40.

```{r hpg-typical}
trial_doses <- function(amt) list(amt = amt, time = 14, ii = 7, addl = 13)
obs_times <- sort(unique(c(seq(0, 280, by = 1), seq(105, 133, by = 0.25))))

sim_typ <- dplyr::bind_rows(lapply(doses, function(a) {
  s <- solve_pk(pk_typ, obs_times, doses = trial_doses(a), covs = covs_median)
  s$treatment <- factor(sprintf("%d mg/wk", a), levels = sprintf("%d mg/wk", doses))
  s
})) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(week = time / 7)
```

```{r fig-tt, fig.cap = "Typical-value total testosterone. Replicates the setting of Bi 2018 Figure 2 (prediction-corrected VPC stratified on dose)."}
ggplot(sim_typ, aes(week, Cc, colour = treatment)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = c(2, 16), linetype = "dashed", colour = "grey40") +
  labs(x = "Study week", y = "Total testosterone (ng/mL)", colour = NULL,
       subtitle = "Dashed lines: first and last active injection") +
  theme_bw()
```

```{r fig-lh, fig.cap = "Typical-value LHRH-stimulated LH30. Replicates the left panel of Bi 2018 Figure 3."}
ggplot(sim_typ, aes(week, lh30, colour = treatment)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = c(2, 16), linetype = "dashed", colour = "grey40") +
  labs(x = "Study week", y = "LH30 (IU/L)", colour = NULL) +
  theme_bw()
```

## Suppression of endogenous testosterone secretion (Figure 4, top panel)

Bi 2018's Figure 4 plots the suppression of endogenous secretion against a
horizontal dashed line representing "basal endogenous testosterone secretion
independent of LH upregulation" -- that is, `b_base`. The structural claim is
that complete loss of LH up-regulation drives the secretion rate down to exactly
`b_base` and no further, and the text states this happened "during week 9 to
week 19 in both the 250 mg and the 500 mg groups".

```{r fig-secretion, fig.cap = "Endogenous testosterone secretion rate. Replicates Bi 2018 Figure 4, top panel."}
b_base <- 6.24
ggplot(sim_typ, aes(week, ksec_te, colour = treatment)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = b_base, linetype = "dashed", colour = "grey30") +
  labs(x = "Study week", y = "Endogenous secretion (mg/day)", colour = NULL,
       subtitle = "Dashed line: b_base, the LH-independent floor") +
  theme_bw()
```

```{r secretion-floor}
floors <- sim_typ |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(min_secretion = min(ksec_te), .groups = "drop") |>
  dplyr::mutate(`Reaches b_base` = abs(min_secretion - b_base) < 0.01)
knitr::kable(floors, digits = 4, caption = "Secretion floor by arm.")

# Structural identity: the high-dose arms must land exactly on b_base, and no arm
# may go below it. A sign error or a mis-transcribed b_base breaks both.
stopifnot(
  all(sim_typ$ksec_te >= b_base - 1e-8),
  floors$`Reaches b_base`[floors$treatment == "250 mg/wk"],
  floors$`Reaches b_base`[floors$treatment == "500 mg/wk"]
)
```

## Recovery timing

Bi 2018 reports when suppression of LH synthesis falls back below 10% after the
last active injection (study week 15): "in week 10 ... in the 500 mg group,
compared after week 5 in the 100 mg group and week 6 in the 250 mg group".

```{r recovery-timing}
recov <- sim_typ |>
  dplyr::filter(week > 15) |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    weeks_after_last_dose = min(week[inh_lh < 0.10]) - 15, .groups = "drop"
  ) |>
  dplyr::mutate(`Bi 2018 Results` = c(5, 6, 10))
knitr::kable(recov, digits = 1,
             caption = "Weeks after the last injection until LH-synthesis suppression falls below 10%.")

# Deterministic typical-value trajectory; the published figures are medians over
# post hoc individuals, so a two-to-three week offset is expected. A bound of 4
# weeks still fails on a mis-transcribed ke0, IC50 or Hill coefficient, each of
# which shifts recovery by many weeks.
stopifnot(all(abs(recov$weeks_after_last_dose - recov$`Bi 2018 Results`) <= 4))
```

# Virtual cohort

A 100-subject-per-arm cohort with the covariate distributions of Table 1.
The Hill coefficient of the LH up-regulation term carries a **proportional**
random effect in the source control stream (`LAM = TVLAM * (1 + ETA(5))`) with a
75.9% coefficient of variation, so roughly 9% of simulated subjects draw a
negative exponent; this is the authors' own parameterisation and is discussed in
the Errata below.

```{r cohort}
n_per_arm <- 100
set.seed(1)
make_cohort <- function(amt, offset) {
  wt <- pmin(pmax(rnorm(n_per_arm, 85, 14), 60.7), 115)
  alb <- pmin(pmax(rnorm(n_per_arm, 45.5, 3.0), 35), 55)
  tibble::tibble(
    id = offset + seq_len(n_per_arm),
    WT_BASE = wt, WT = wt + rnorm(n_per_arm, 2.90, 1.5),
    ALB_BASE = alb, ALB = alb + rnorm(n_per_arm, -2.0, 1.5),
    T4 = pmin(pmax(rnorm(n_per_arm, 7.40, 1.2), 4.5), 12),
    treatment = factor(sprintf("%d mg/wk", amt), levels = sprintf("%d mg/wk", doses))
  )
}
cohort <- dplyr::bind_rows(lapply(seq_along(doses), function(i) {
  make_cohort(doses[i], (i - 1L) * n_per_arm)
}))

vpc_times <- sort(unique(c(seq(0, 280, by = 3.5), seq(105, 133, by = 1))))
events <- cohort |>
  dplyr::select(id, treatment) |>
  tidyr::crossing(time = vpc_times) |>
  dplyr::mutate(evid = 0L, amt = NA_real_, cmt = "central", dvid = 1L)
dose_rows <- cohort |>
  dplyr::select(id, treatment) |>
  tidyr::crossing(time = 14 + 7 * (0:13)) |>
  dplyr::mutate(evid = 1L, cmt = "depot", dvid = NA_integer_) |>
  dplyr::mutate(amt = doses[match(treatment, levels(treatment))])
events <- dplyr::bind_rows(events, dose_rows) |>
  dplyr::left_join(dplyr::select(cohort, -treatment), by = "id") |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim_cohort <- rxode2::rxSolve(
  uis[[1]], events, returnType = "data.frame", useLinCmt = FALSE,
  keep = c("treatment")
) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(week = time / 7)
cat("cohort rows:", nrow(sim_cohort), " subjects:", dplyr::n_distinct(sim_cohort$id), "\n")
```

```{r fig-vpc, fig.cap = "Simulated total testosterone percentiles by arm (2.5th, 50th, 97.5th). Compare Bi 2018 Figure 2."}
pctl <- sim_cohort |>
  dplyr::group_by(treatment, week) |>
  dplyr::summarise(
    lo = quantile(ipredSim, 0.025), md = median(ipredSim),
    hi = quantile(ipredSim, 0.975), .groups = "drop"
  )
ggplot(pctl, aes(week)) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = treatment), alpha = 0.2) +
  geom_line(aes(y = md, colour = treatment), linewidth = 0.8) +
  labs(x = "Study week", y = "Total testosterone (ng/mL)",
       colour = NULL, fill = NULL) +
  theme_bw()
```

```{r fig-vpc-lh, fig.cap = "Simulated LH30 percentiles by arm. Compare Bi 2018 Figure 3, left panel."}
pctl_lh <- sim_cohort |>
  dplyr::group_by(treatment, week) |>
  dplyr::summarise(
    lo = quantile(lh30, 0.025), md = median(lh30),
    hi = quantile(lh30, 0.975), .groups = "drop"
  )
ggplot(pctl_lh, aes(week)) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = treatment), alpha = 0.2) +
  geom_line(aes(y = md, colour = treatment), linewidth = 0.8) +
  labs(x = "Study week", y = "LH30 (IU/L)", colour = NULL, fill = NULL) +
  theme_bw()
```

Bi 2018 reports that LH suppression was severe and consistent in the two high
dose arms but variable in the 100 mg arm: "the suppression of LH was more severe
in the 250 mg and the 500 mg groups, whereas the suppression is variable in the
100 mg group."

```{r lh-variability}
nadir <- sim_cohort |>
  dplyr::filter(week >= 7, week <= 16) |>
  dplyr::group_by(treatment, id) |>
  dplyr::summarise(max_inh = max(inh_lh), .groups = "drop") |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    `Median LH-synthesis suppression` = median(max_inh),
    `5th percentile` = quantile(max_inh, 0.05),
    `Interquartile width` = IQR(max_inh), .groups = "drop"
  )
knitr::kable(nadir, digits = 3,
             caption = "Peak LH-synthesis suppression across weeks 7-16, by arm.")

# Cohort-derived, so assert only the ORDERING claim the paper makes, with
# headroom: the two high-dose arms are near-complete and much less dispersed
# than the 100 mg arm.
stopifnot(
  nadir$`Median LH-synthesis suppression`[nadir$treatment == "250 mg/wk"] > 0.9,
  nadir$`Median LH-synthesis suppression`[nadir$treatment == "500 mg/wk"] > 0.9,
  nadir$`Interquartile width`[nadir$treatment == "100 mg/wk"] >
    nadir$`Interquartile width`[nadir$treatment == "500 mg/wk"]
)
```

# Suppression of spermatogenesis

The sperm model consumes `CAV`, the average total testosterone concentration
over the preceding 18 weeks. Bi 2018 computed it from the individual post hoc PK
predictions; here it is computed from the typical-value PK simulation above,
using the paper's own rule that observations earlier than 18 weeks average from
the start of the study.

```{r cav}
tau_window <- 18 * 7 # days

cav18 <- function(time, conc) {
  auc <- c(0, cumsum(diff(time) * (head(conc, -1) + tail(conc, -1)) / 2))
  lower <- pmax(0, time - tau_window)
  auc_lower <- approx(time, auc, xout = lower)$y
  width <- pmax(time - lower, .Machine$double.eps)
  out <- (auc - auc_lower) / width
  out[time == 0] <- conc[time == 0] # limit of AUC(t)/t as t -> 0
  out
}

cav_df <- sim_typ |>
  dplyr::group_by(treatment) |>
  dplyr::arrange(time, .by_group = TRUE) |>
  dplyr::mutate(CAV = cav18(time, Cc)) |>
  dplyr::ungroup()
```

```{r fig-cav, fig.cap = "The 18-week rolling average testosterone concentration that drives the spermatogenesis model."}
ggplot(cav_df, aes(week, CAV, colour = treatment)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 8.68, linetype = "dashed", colour = "grey30") +
  labs(x = "Study week", y = "CAV: 18-week average tT (ng/mL)", colour = NULL,
       subtitle = "Dashed line: Cavg50 = 8.68 ng/mL at 85 kg") +
  theme_bw()
```

The pre-treatment value of `CAV` is not zero -- the analyte is endogenous -- and
the model's unsuppressed initial condition depends on that baseline sitting well
below `Cavg50`. That is a sharper constraint than it looks, because the Hill
coefficient is 11.3: production at baseline must be essentially unsuppressed for
`sperm(0) = kin/kout` to be a self-consistent initial condition.

```{r sperm-sim}
sim_sperm <- dplyr::bind_rows(lapply(levels(cav_df$treatment), function(tr) {
  d <- dplyr::filter(cav_df, treatment == tr)
  ev <- data.frame(id = 1L, time = d$time, evid = 0L, amt = NA_real_,
                   dvid = 1L, CAV = d$CAV, WT_BASE = 85)
  s <- rxode2::rxSolve(sp_typ, ev, returnType = "data.frame", useLinCmt = FALSE)
  s$treatment <- factor(tr, levels = levels(cav_df$treatment))
  s
})) |>
  dplyr::mutate(week = time / 7)

baseline_sperm <- 6.17 / 0.0696
sim_sperm <- dplyr::mutate(
  sim_sperm, suppression = 100 * (1 - sperm / baseline_sperm)
)

# Baseline must be essentially unsuppressed: this is what fixes the weight
# reference at 85 kg rather than the 70 kg printed in the deposited stream.
base_supp <- sim_sperm |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(baseline_suppression = suppression[which.min(time)], .groups = "drop")
knitr::kable(base_supp, digits = 3,
             caption = "Suppression of sperm production at time zero (should be near zero).")
stopifnot(all(abs(base_supp$baseline_suppression) < 5))
```

```{r fig-sperm, fig.cap = "Typical-value sperm count. Replicates Bi 2018 Figure 3 (right panel) and Figure 4 (bottom panel)."}
ggplot(sim_sperm, aes(week, sperm, colour = treatment)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = c(2, 16), linetype = "dashed", colour = "grey40") +
  labs(x = "Study week", y = "Sperm count", colour = NULL) +
  theme_bw()
```

```{r sperm-claims}
at_week <- function(tr, w) {
  d <- dplyr::filter(sim_sperm, treatment == tr)
  approx(d$week, d$suppression, xout = w)$y
}
sperm_claims <- tibble::tribble(
  ~Claim, ~Week, ~Arm, ~`Bi 2018`,
  "Sperm count completely suppressed at the end of dosing", 15, "250 mg/wk", 98.2,
  "Sperm count completely suppressed at the end of dosing", 15, "500 mg/wk", 94.9,
  "Partial recovery by week 28", 28, "100 mg/wk", 26.6,
  "Partial recovery by week 28", 28, "250 mg/wk", 48.4,
  "Partial recovery by week 28", 28, "500 mg/wk", 66.9
) |>
  dplyr::mutate(
    Simulated = mapply(at_week, Arm, Week),
    Deviation = Week == 28
  )
knitr::kable(sperm_claims, digits = 1,
             caption = paste("Suppression of sperm count (%) against Bi 2018's",
                             "post hoc medians. Rows flagged Deviation are not gated;",
                             "see the Errata."))

# The end-of-dosing claim is the one the paper states most confidently and the
# regime where the response is saturated, so it is gated.
stopifnot(
  all(sperm_claims$Simulated[sperm_claims$Week == 15] > 90),
  # Dose ordering of the week-28 recovery is preserved even though the
  # magnitudes are not.
  at_week("500 mg/wk", 28) > at_week("250 mg/wk", 28),
  at_week("250 mg/wk", 28) > at_week("100 mg/wk", 28)
)
```

At the end of dosing, where the paper's estimates are precise ("suppression at
week 15 and week 21 were precisely estimated"), the model reproduces the
published values closely. The week-28 recovery values are systematically higher
in the typical-value simulation than in the published post hoc medians. This is
expected and the paper anticipates it: recovery "was estimated with wide CIs",
the inter-individual variability on the logit-Emax parameter is 234%, and seven
of 29 subjects missed the final collection. A typical-value trajectory is not
the median of a cohort that variable, and the paper itself flags the
end-of-study estimate as possibly biased.

# Assumptions and deviations

### Values taken from the supplementary control streams rather than the article

The main article prints the differential equations but not the covariate
functional forms, their reference and centring constants, or the residual-error
parameterisation. All of these come from the NONMEM control streams in
Supplementary Material S6, which were retrieved from the EuropePMC
supplementary-file endpoint for PMC5915615. In particular:

* The **residual-error scale differs between layers, and the tables' labels are
  wrong in one direction**. The PK stream builds
  `W = SQRT(IPRED*IPRED*THETA(8) + THETA(9))` with `$SIGMA 1 FIX`, so Table 2's
  `sigma^2 proportional` (0.057) and `sigma^2 additive` (0.261) really are
  variances and the model file stores their square roots (0.239 and 0.511). The
  LH and sperm streams instead set `W = THETA(6)` directly, so Table 3's
  `sigma^2 additive` entries (0.397 and 0.429) are **standard deviations**
  despite the label. Reading either table the other way would misstate the
  residual magnitude by a factor of roughly two.
* Both PD endpoints are fitted on a **log(state + 1)** scale
  (`IPRED = LOG(A(4) + 1)`). The `+1` offset is load-bearing rather than
  cosmetic, because sperm counts and LH30 both approach zero under complete
  suppression in the high-dose arms.
* The `A_0, TC` row of Table 2 is a **concentration**, not an amount: the stream
  computes `INT = THETA(10) * EXP(ETA(6))` and then `A_0(2) = INT * V`. Its
  value (6.24 ng/mL) coincidentally equals `b_base` (6.24 mg/day), which are
  different quantities with different units, different RSEs and different
  confidence intervals.
* The `$THETA` comment labels in the PK stream are **stale and shifted** from
  `THETA(8)` onwards: the entry commented `[K1 500mg]` is the proportional
  residual variance, `[proportional error]` is the additive residual variance,
  and `[additive error]` is the baseline concentration. Each was mapped by how
  the stream *uses* it, not by its comment, and each maps onto the matching
  Table 2 row.

### The deposited control streams are not uniformly final

The PK and LH streams carry final estimates -- 14 of 15 PK `$THETA` values and
all 6 PK `$OMEGA` values reproduce Table 2 exactly, and every LH `$THETA` and
`$OMEGA` reproduces Table 3. The **sperm-count stream does not**: none of its
seven `$THETA` values matches Table 3 (for example `0.1` against a final
`-1.27`, an obvious starting value), and its `$OMEGA` entries do not reproduce
the published IIV percentages either. All parameter *values* in both model files
therefore come from Tables 2 and 3; the streams are used only for model
*structure*, functional forms and constants.

Two consequences:

1. **`h Balbumin-V`.** Table 2 gives the final estimate as -1.55 (bootstrap
   median -1.84); the PK stream's `$THETA(15)` is -1.82851. The model file uses
   the published -1.55. The parameter has 58.7% RSE, and two other Table 2
   parameters (`Lambda` and the additive residual) show a comparably large gap
   between final estimate and bootstrap median, so a final value well away from
   the bootstrap median is not anomalous here. The albumin effect is one the
   paper explicitly calls not clinically important. Using -1.83 instead changes
   the typical volume by under 2% at plausible albumin values.
2. **The weight reference for `Cavg50`.** The deposited sperm stream prints
   `(WT/70)`, but that stream carries starting values throughout. The model file
   uses **85 kg**, on three grounds: it is the typical weight Bi 2018 names in
   the sentence reporting this very effect; it is the reference used by every
   other covariate in the paper (85 kg in the PK layer, 84.70 kg in the LH
   layer); and it is the only choice consistent with the model's own initial
   condition. With a 70 kg reference the typical `Cavg50` at 85 kg falls to
   6.78 ng/mL, which -- against a baseline `CAV` near 6.2 ng/mL and a Hill
   coefficient of 11.3 -- predicts roughly 28% suppression of sperm production
   in untreated healthy men, contradicting `A_0 = KIN/KOUT`. The check in the
   spermatogenesis section above gates exactly this. It also reproduces the
   paper's week-28 recovery figures far better (18% versus 78% against a
   published 26.6% for the 100 mg arm).

### Coupling the sequentially-fitted layers

Bi 2018 fitted testosterone PK with LH supplied as a data column and fitted LH
with the PK parameters fixed per subject. `Bi_2018_testosteroneCypionate` closes
that loop, using the model's own `lh30` state where the paper used its post hoc
prediction. Two small consequences follow. The stream's branch
`IF (LHORI .EQ. 0) k1 = BASE` is replaced by a floor of 1e-8 on the LH driver,
which is numerically equivalent and keeps the power term finite. And the LH
stream's hard-coded initial testosterone concentration of 6.5 ng/mL
(`A_0(2) = 6.5*V`) is replaced by the PK model's own estimated baseline of
6.24 ng/mL, so the two layers share one baseline instead of two.

### The Hill coefficient of the LH up-regulation term can go negative

The PK stream codes `LAM = TVLAM * (1 + ETA(5))` -- a proportional, not
log-normal, random effect -- with `$OMEGA` 0.576, i.e. a 75.9% CV. Roughly 9% of
simulated subjects therefore draw a negative Hill exponent, which inverts the
direction of LH's effect on testosterone secretion for those subjects. This is
the authors' own parameterisation and is reproduced faithfully rather than
silently regularised; the term stays bounded in (0, 1) either way, so it
produces no numerical failure. Users running typical-value simulations
(`rxode2::zeroRe()`) are unaffected. Note the contrast with the LH and sperm
layers, where the same Hill coefficient carries a conventional log-normal effect
(`LAM = TVLAM * EXP(ETA)`).

### Covariates

* **`T4` is total thyroxine, not thyroxine-binding globulin.** Bi 2018 Table 1
  tabulates thyroxine-binding globulin (group medians 19.5 / 18.5 / 21 ug/mL),
  but the covariate in the LH control stream is `THYROXIN` referenced to 7.40,
  and the supplementary LH dataset carries `thyroxin = 7.4` for its example
  subject -- consistent with total T4 in ug/dL and inconsistent with TBG. The
  baseline distribution of this covariate is therefore **not published**, and
  the virtual cohort above assumes a modest spread around the 7.40 reference.
  The paper reports that this covariate made no visible difference to simulated
  LH suppression.
* **The virtual cohort's covariate distributions are assumed.** Table 1 gives
  per-arm medians and ranges but no standard deviations and no correlation
  structure, so normal distributions truncated to the published ranges are used.
  The weight-change and albumin-change distributions are centred on the model's
  own centring constants (+2.90 kg and -0.2 g/dL), which are the cohort medians
  of those changes, with an assumed spread.
* **Serum albumin is held in canonical SI g/L** per the nlmixr2lib covariate
  register, and converted inline to the g/dL scale on which Bi 2018 calibrated
  its coefficients.

### Convention deviations

`checkModelConventions()` reports one warning on
`Bi_2018_testosteroneCypionate_sperm`: the single observation variable
`logSperm` is not a canonical output name. Sperm count is a genuinely
paper-specific endpoint with no canonical compartment in nlmixr2lib and, as the
first paper in the library to carry it, does not meet the two-paper bar for
minting one; the state is declared through `paper_specific_compartments`
instead. The remaining message on both models is informational: `units$dosing`
is mg while `units$concentration` is ng/mL, which is correct here because the
volumes are expressed in kilolitres, so `mg / kL = ug/L = ng/mL` with no
conversion factor.

### Not extracted

The sperm-motility model (Supplementary Table S1) and the LH60 and LH120 models
are described by Bi 2018 as sensitivity analyses that "had similar parameter
estimation and comparable performance", so they are excluded per nlmixr2lib's
replicate-the-author's-structure policy, which excludes robustness checks the
authors did not report as final.

# Session info

```{r session}
sessionInfo()
```
