---
title: "Molnupiravir (Bihorel 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Molnupiravir (Bihorel 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
rxode2::rxSetSeed(20231205)
set.seed(20231205)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Bihorel_2023_molnupiravir"))$reference`
* Description: `r rxode2::rxode(readModelDb("Bihorel_2023_molnupiravir"))$description`
* Article: <https://doi.org/10.1002/psp4.13031>

Molnupiravir (MK-4482, EIDD-2801) is an orally administered ribonucleoside
prodrug. It is essentially undetectable in plasma because it is hydrolysed to
beta-D-N4-hydroxycytidine (NHC) during absorption and first pass, so NHC is the
analyte that was measured and modelled. Bihorel 2023 converted all dose amounts
to molar units with the molnupiravir molecular mass (329.31 Da) and all
concentrations with the NHC molecular mass (259.2 Da); the packaged model keeps
those units, so an 800 mg dose enters as
`800 / 329.31 * 1e6 = `r format(round(800 / 329.31 * 1e6), big.mark = ",")`` nmol
and `Cc` is in nmol/L.

# Population

The analysis pooled 4202 plasma NHC concentrations from 1207 participants in
four randomised, double-blind, placebo-controlled trials (Bihorel 2023
Table 1): the phase I trial MK-4482-004 in 100 healthy adults, the phase IIa
trial MK-4482-006 in 66 non-hospitalised participants with COVID-19, the
phase II trial MOVe-IN (MK-4482-001) in 196 hospitalised participants, and the
phase II/III trial MOVe-OUT (MK-4482-002) in 845 non-hospitalised participants.
Overall 48.3% were women, median (range) age was 46 (18-91) years, median body
weight 85 (36.1-172) kg and median BMI 30.4 (14.3-68.6) kg/m^2. Two thirds
(66.7%) were White. Renal impairment was common (48.1% mild, 7.0% moderate);
hepatic impairment, graded with a modified Child-Pugh score approximated from
bilirubin and albumin, was rare (5.0% mild, 0.2% moderate).

Sampling was highly unbalanced: healthy participants contributed 7-26 samples
each while participants with COVID-19 contributed 1-5, so 72.8% of the pooled
population contributed at most two samples. That imbalance is why the model
carries two proportional residual-error magnitudes, selected by
`STUDY_MOV_PHASE23`.

The same information is available programmatically via
`readModelDb("Bihorel_2023_molnupiravir")()$population`.

# Source trace

Every value below is also recorded as an in-file comment beside its `ini()`
entry in `inst/modeldb/specificDrugs/Bihorel_2023_molnupiravir.R`.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` | 0.797 1/h | Table 2, row "k a / First-order absorption rate constant, 1/h" (RSE 2.57%) |
| `lmtt` | 0.435 h | Table 2, row "MTT / Mean absorption transit time, h" (RSE 5.39%) |
| `nn` | 7.84 | Table 2, row "NN / Number of transit compartments" (RSE 16.5%) |
| `e_highfat_mtt` | 4.22 | Table 2, row "Proportional shift due to high-fat meal" (RSE 6.29%); Results: "422% increase in MTT" |
| `e_solution_mtt` | -0.616 | Table 2, row "Proportional shift in oral solution" (RSE 5.49%); Results: "61.6% decrease in MTT" |
| `lfdepot` | fixed at 1 | Table 2, row "F1 / Relative bioavailability" = 1.00, %RSE column reads "Fixed" |
| `lcl` | 70.6 L/h at 80 kg | Table 2, row "CL/F / Apparent central clearance in 80kg participants, L/h" (RSE 1.97%) |
| `e_wt_cl` | 0.412 | Table 2, row "Power of body weight effect" (RSE 14.0%) |
| `lvc` | 63.9 L at BMI 28, male | Table 2, row "V C /F / Apparent central volume in 28kg/m 2 BMI male participants, L" (RSE 5.07%) |
| `e_bmi_vc` | 0.997 | Table 2, row "Power of BMI effect" (RSE 13.1%) |
| `e_sexf_vc` | -0.330 | Table 2, row "Proportional shift in female participants" (RSE 11.8%) |
| `lq` | 2.99 L/h | Table 2, row "Q/F / Apparent distribution clearance, L/h" (RSE 5.70%) |
| `lvp` | 68.3 L | Table 2, row "V P /F / Apparent peripheral volume, L" (RSE 14.6%) |
| `etalcl` | 0.172573 | Table 2, CL/F variability 43.4 %CV; `log(1 + 0.434^2)` |
| `etalvc` | 0.333371 | Table 2, V C /F variability 62.9 %CV; `log(1 + 0.629^2)` |
| `etaiov_mtt_1`, `etaiov_mtt_2` | 0.147049 | Table 2, row "IOV in MTT" 39.8 %CV; `log(1 + 0.398^2)`; two occasions per the Table 2 note |
| `propSdPhase1` | `sqrt(0.0652)` = 0.255 | Table 2, "Residual variability / Phase I trials" = 0.0652, 25.5 %CV |
| `propSdPhase23` | `sqrt(0.247)` = 0.497 | Table 2, "Residual variability / Phase II/III trials" = 0.247, 49.7 %CV |
| Absorption structure (transit chain into depot, first-order `ka` into central) | n/a | Figure 1a (stages 1 and 3); Results, "Base model refinement using phase III data" |
| Two-compartment linear disposition | n/a | Abstract; Results, "Base structural model development" |
| Dose / concentration molar conversion | 329.31 and 259.2 Da | Methods, "Bioanalytical methods" |
| Reference individual for the typical profile | 78 kg, BMI 28, man, non-hospitalised | Discussion, paragraph beginning "Simulation of the typical PK profile" |
| Published typical Cmax / Tmax / t1/2 | 11,400 nmol/L, ~1.5 h, 0.6 and 16.5 h | Discussion, same paragraph |
| AUC(0-12) clinical comparability bounds | 23,800-68,000 nmol*h/L | Methods, "Assessment of clinical relevance ..." |
| Reference AUC(0-12) used in the NCA comparison | 34,000 nmol*h/L | Derived from the line above by inverting the printed bounds: 23,800 / 0.7 = 68,000 / 2.0 = 34,000 |

# Virtual cohort

Original observed data are not publicly available. The simulations below use
virtual populations whose covariate distributions approximate the published
demographics of Table 1.

```{r cohort}
DOSE_800MG_NMOL <- 800 / 329.31 * 1e6  # Methods, "Bioanalytical methods"

# Bihorel 2023 Discussion reference individual: a non-hospitalised man of
# 78 kg and BMI 28 kg/m^2 given 800 mg q12h for 5 days (10 doses). Times are
# re-based below so that the LAST dose sits at t = 0, which makes Tmax and the
# terminal half-life unambiguous for PKNCA.
LAST_DOSE_TIME <- 108

ev_typical <- rxode2::et(amt = DOSE_800MG_NMOL, cmt = "depot", ii = 12, addl = 9) |>
  # Observe the ODE state, never the algebraic observable `Cc`.
  rxode2::et(seq(LAST_DOSE_TIME, 240, by = 0.05), cmt = "central") |>
  as.data.frame() |>
  mutate(
    WT = 78, BMI = 28, SEXF = 0,
    FED_HIGHFAT = 0, FORM_SOLUTION = 0,
    OCC = 1, STUDY_MOV_PHASE23 = 0,
    treatment = "Reference individual"
  )
```

The food-effect and formulation arms below are single 800 mg doses in the same
reference individual, differing only in `FED_HIGHFAT` / `FORM_SOLUTION`. Because
both covariates act solely on the mean transit time and relative bioavailability
is fixed at 1, these arms isolate a pure absorption-rate effect.

```{r cohort-food}
make_single_dose_arm <- function(label, fed_highfat, form_solution, id) {
  rxode2::et(amt = DOSE_800MG_NMOL, cmt = "depot") |>
    rxode2::et(seq(0, 72, by = 0.05), cmt = "central") |>
    as.data.frame() |>
    mutate(
      id = id, WT = 78, BMI = 28, SEXF = 0,
      FED_HIGHFAT = fed_highfat, FORM_SOLUTION = form_solution,
      OCC = 1, STUDY_MOV_PHASE23 = 0,
      treatment = label
    )
}

ev_food <- bind_rows(
  make_single_dose_arm("Capsule, fasted",       0, 0, 1L),
  make_single_dose_arm("Capsule, high-fat meal", 1, 0, 2L),
  make_single_dose_arm("Oral solution, fasted",  0, 1, 3L)
)
stopifnot(!anyDuplicated(unique(ev_food[, c("id", "time", "evid")])))
```

The stochastic cohort mimics the non-hospitalised phase II/III population: 200
participants (the per-arm cap), capsules only, with `FED_HIGHFAT` drawn from the
25% mixture fraction Bihorel 2023 fixed for participants whose food status was
not collected.

```{r cohort-iiv}
N_COHORT <- 200L

# Body weight and BMI are strongly correlated at fixed height; the paper reports
# each marginally (Table 1) but not their correlation, so a bivariate normal with
# rho = 0.8 is assumed and both margins are truncated to the observed ranges.
# See "Assumptions and deviations".
rho <- 0.8
z1 <- rnorm(N_COHORT)
z2 <- rho * z1 + sqrt(1 - rho^2) * rnorm(N_COHORT)
cohort_cov <- tibble(
  id  = seq_len(N_COHORT),
  WT  = pmin(pmax(85.8 + 18.4 * z1, 36.1), 172),   # Table 1 overall mean (SD), range
  BMI = pmin(pmax(30.4 +  6.1 * z2, 14.3), 68.6),  # Table 1 overall mean (SD), range
  SEXF = rbinom(N_COHORT, 1, 0.483),                   # Table 1: 48.3% women
  FED_HIGHFAT = rbinom(N_COHORT, 1, 0.25),             # Results: mixture fraction fixed to 25%
  FORM_SOLUTION = 0,                                   # Table 1: capsules only in phase II/III
  OCC = 1,
  STUDY_MOV_PHASE23 = 1
)

# Coarse grid over the accumulation phase, fine grid over the final interval
# where the NCA is taken.
obs_times <- sort(unique(c(seq(0, 108, by = 0.5), seq(108, 120, by = 0.05))))

ev_template <- rxode2::et(amt = DOSE_800MG_NMOL, cmt = "depot", ii = 12, addl = 9) |>
  rxode2::et(obs_times, cmt = "central") |>
  as.data.frame()
# et() may emit its own single-subject `id`; drop it so `cohort_cov$id` is the
# only subject key.
ev_template$id <- NULL

ev_cohort <- tidyr::crossing(cohort_cov, ev_template) |>
  arrange(id, time, desc(evid))
stopifnot(!anyDuplicated(unique(ev_cohort[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod <- readModelDb("Bihorel_2023_molnupiravir")
mod_typical <- rxode2::zeroRe(mod)

sim_typical <- rxode2::rxSolve(
  mod_typical, events = ev_typical, keep = "treatment",
  omega = NA, sigma = NA, returnType = "data.frame"
)

sim_food <- rxode2::rxSolve(
  mod_typical, events = ev_food, keep = "treatment",
  omega = NA, sigma = NA, returnType = "data.frame"
)

sim_cohort <- rxode2::rxSolve(
  mod, events = ev_cohort, keep = c("WT", "BMI", "SEXF", "FED_HIGHFAT"),
  returnType = "data.frame"
)
```

`Cc` is the individual prediction; residual error appears in the `sim` column.
The NCA below is deliberately run on `Cc`, because the published values being
reproduced are model predictions rather than observations.

# Replicate published results

## Typical steady-state profile (Discussion)

Bihorel 2023 report that, for the reference individual, "NHC concentration is
expected to peak at 11,400 nmol/L ~1.5 h after the last dose, with half-lives of
0.6 and 16.5 h for the first and second phases of disposition."

```{r figure-typical}
sim_typical |>
  mutate(tad = time - LAST_DOSE_TIME) |>
  filter(tad >= 0, tad <= 48) |>
  ggplot(aes(tad, Cc)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 11400, linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = 1.5, linetype = "dashed", colour = "grey40") +
  scale_y_log10() +
  labs(
    x = "Time after the last dose (h)", y = "Plasma NHC (nmol/L)",
    title = "Typical NHC profile after the last of ten 800 mg q12h doses",
    caption = paste(
      "Reference individual: non-hospitalised man, 78 kg, BMI 28 kg/m^2.",
      "Dashed lines mark the published Cmax of 11,400 nmol/L at ~1.5 h",
      "(Bihorel 2023, Discussion)."
    )
  )
```

The biexponential decline is a direct consequence of the disposition
micro-constants, which can be checked in closed form against the two published
half-lives without any simulation at all.

```{r halflife-identity}
cl_ref  <- 70.6 * (78 / 80)^0.412
vc_ref  <- 63.9
kel <- cl_ref / vc_ref; k12 <- 2.99 / vc_ref; k21 <- 2.99 / 68.3
b <- kel + k12 + k21
d <- sqrt(b^2 - 4 * kel * k21)
hl <- log(2) / c(alpha = (b + d) / 2, beta = (b - d) / 2)

tibble(
  Phase = c("First (alpha)", "Second (beta)"),
  Model = round(unname(hl), 2),
  Published = c(0.6, 16.5)
) |>
  rename("Half-life, model (h)" = Model, "Half-life, published (h)" = Published) |>
  knitr::kable(caption = "Disposition half-lives from the model's micro-constants vs Bihorel 2023 Discussion.")

# Deterministic closed-form identity: both sides use the same parameters, so a
# tight bound is correct here (this is numerical error, not cohort variability).
stopifnot(abs(hl[["alpha"]] - 0.6) < 0.05, abs(hl[["beta"]] - 16.5) < 0.2)
```

## Food and formulation act on absorption rate only

```{r figure-food}
sim_food |>
  filter(time <= 12) |>
  ggplot(aes(time, Cc, colour = treatment)) +
  geom_line(linewidth = 0.8) +
  labs(
    x = "Time (h)", y = "Plasma NHC (nmol/L)", colour = NULL,
    title = "Single 800 mg dose: effect of a high-fat meal and of the oral solution",
    caption = paste(
      "A high-fat meal raises the mean transit time 422% and the oral solution",
      "lowers it 61.6% (Bihorel 2023 Table 2); neither changes the extent of",
      "absorption."
    )
  ) +
  theme(legend.position = "bottom")
```

Because `FED_HIGHFAT` and `FORM_SOLUTION` act only on the mean transit time and
`F1` is fixed at 1, total exposure must be identical across the three arms. That
is an internal identity of the model rather than a cohort statistic, so it is
asserted tightly.

```{r food-auc-identity}
auc_by_arm <- sim_food |>
  filter(!is.na(Cc)) |>
  group_by(treatment) |>
  summarise(
    auc72 = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    cmax = max(Cc), tmax = time[which.max(Cc)], .groups = "drop"
  )

auc_by_arm |>
  rename(
    "Arm" = treatment, "AUC(0-72) (nmol*h/L)" = auc72,
    "Cmax (nmol/L)" = cmax, "Tmax (h)" = tmax
  ) |>
  knitr::kable(digits = c(0, 0, 0, 2),
               caption = "Single 800 mg dose by food status and formulation.")

# AUC must be invariant to MTT; only rate changes.
stopifnot(diff(range(auc_by_arm$auc72)) / mean(auc_by_arm$auc72) < 1e-3)
# The high-fat meal must delay and blunt the peak; the solution must sharpen it.
tmax_fasted <- auc_by_arm$tmax[auc_by_arm$treatment == "Capsule, fasted"]
stopifnot(
  auc_by_arm$tmax[auc_by_arm$treatment == "Capsule, high-fat meal"] > tmax_fasted,
  auc_by_arm$tmax[auc_by_arm$treatment == "Oral solution, fasted"] < tmax_fasted
)
```

## Stochastic cohort (Figure 2 / Figure S5 in spirit)

```{r figure-vpc}
sim_cohort |>
  filter(time >= 96, time <= 120) |>
  group_by(time) |>
  summarise(
    Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time - 96, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(
    x = "Time after the ninth dose (h)", y = "Plasma NHC (nmol/L)",
    title = "Simulated 5th / 50th / 95th percentiles over the last two dosing intervals",
    caption = paste(
      "200 virtual non-hospitalised participants, 800 mg q12h for 5 days.",
      "Comparable in construction to the prediction-corrected VPC of",
      "Bihorel 2023 Figure 2."
    )
  )
```

At steady state the model implies `AUC(0-12) = dose / (CL/F)` exactly, which
gives a per-subject identity that does not depend on the drawn cohort.

```{r auc-identity}
auc_tau <- sim_cohort |>
  filter(time >= 108, !is.na(Cc)) |>
  group_by(id) |>
  arrange(time, .by_group = TRUE) |>
  summarise(
    auc12 = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    WT = first(WT), .groups = "drop"
  ) |>
  # CL/F per subject, recovered from the identity AUC = dose / CL.
  mutate(cl_implied = DOSE_800MG_NMOL / auc12)

summary_tbl <- tibble(
  Quantity = c(
    "Median AUC(0-12) (nmol*h/L)",
    "5th percentile AUC(0-12)",
    "95th percentile AUC(0-12)",
    "Median CL/F implied by AUC (L/h)"
  ),
  Value = c(
    median(auc_tau$auc12), quantile(auc_tau$auc12, 0.05),
    quantile(auc_tau$auc12, 0.95), median(auc_tau$cl_implied)
  )
)
summary_tbl |>
  knitr::kable(digits = 0,
               caption = "Simulated steady-state exposure, 800 mg q12h, 200 virtual participants.")

# Structural gates. These are centre / robust-quantile statements, not extremes,
# so they hold for any cohort this model can produce.
#  * The typical CL/F must sit near the published 70.6 L/h at 80 kg. The cohort
#    median weight is above 80 kg, so the median implied CL/F runs slightly high.
stopifnot(abs(median(auc_tau$cl_implied) - 70.6) / 70.6 < 0.25)
#  * The median AUC(0-12) must fall inside the paper's clinical comparability
#    bounds of 23,800-68,000 nmol*h/L (Methods).
stopifnot(median(auc_tau$auc12) > 23800, median(auc_tau$auc12) < 68000)
#  * AUC must fall with body weight (power -0.412 on CL/F). Compare the medians
#    of the lightest and heaviest weight terciles rather than any single subject.
terciles <- cut(auc_tau$WT, quantile(auc_tau$WT, c(0, 1/3, 2/3, 1)),
                include.lowest = TRUE, labels = c("light", "mid", "heavy"))
med_by_tercile <- tapply(auc_tau$auc12, terciles, median)
stopifnot(med_by_tercile[["light"]] > med_by_tercile[["heavy"]])
```

# PKNCA validation

```{r pknca}
# Re-base time so the last dose sits at t = 0. `ev_typical` is a single
# deterministic subject, so rxSolve() returns no `id` column; PKNCA needs one,
# and adding it here is safe precisely because there is definitionally one
# subject in this arm.
sim_nca <- sim_typical |>
  filter(!is.na(Cc)) |>
  mutate(id = 1L, time = time - LAST_DOSE_TIME) |>
  filter(time >= 0) |>
  select(id, time, Cc, treatment)

# Guarantee a time-zero record per (id, treatment). Only `!is.na(Cc)` is used as
# a filter above, per the PKNCA recipe.
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- sim_nca |>
  distinct(id, treatment) |>
  mutate(time = 0, amt = DOSE_800MG_NMOL)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start     = c(0, 12),
  end       = c(12, Inf),
  cmax      = c(TRUE, FALSE),
  tmax      = c(TRUE, FALSE),
  auclast   = c(TRUE, FALSE),
  half.life = c(FALSE, TRUE)
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

## Comparison against published values

```{r nca_compare}
# PKNCA returns `tmax` for BOTH intervals, because it computes tmax internally
# as a dependency of half.life on the terminal interval -- and there the maximum
# sits at the interval start, giving tmax = 0. Averaging that with the real
# within-dose tmax of 1.5 h would report a spurious 0.75 h, so each parameter is
# taken from the interval it was actually requested on: peak and exposure from
# the 0-12 h dosing interval, half-life from the 12 h-onward terminal interval.
nca_long <- as.data.frame(nca_res$result) |>
  filter(
    (PPTESTCD %in% c("cmax", "tmax", "auclast") & start == 0) |
      (PPTESTCD == "half.life" & start == 12)
  ) |>
  select(treatment, PPTESTCD, PPORRES)

published <- tibble::tribble(
  ~treatment,             ~cmax, ~tmax, ~half.life, ~auclast,
  "Reference individual", 11400, 1.5,   16.5,       34000
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_long,
  reference     = published,
  by            = "treatment",
  units         = c(cmax = "nmol/L", tmax = "h", half.life = "h", auclast = "nmol*h/L"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated vs published typical values for the Bihorel 2023 reference",
    "individual (Discussion). * differs from the reference by >20%."
  )
)
```

The reference AUC(0-12) of 34,000 nmol*h/L is not printed as such; it is
recovered from the paper's own printed correspondence in Methods, "the 0.7-2.0
range corresponds to plasma NHC AUC(0-12) of ~23,800-68,000 nmol*h/L with
molnupiravir 800 mg q12h". Both endpoints invert to the same reference value
(23,800 / 0.7 = 34,000 and 68,000 / 2.0 = 34,000), so the derivation is
arithmetic on two printed numbers that agree, not a digitised or assumed value.

The simulated Cmax, Tmax, terminal half-life and AUC(0-12) all reproduce the
published typical values well within the 20% flagging threshold, so the
transcription of the absorption chain, the disposition micro-constants and the
molar dose conversion are all confirmed simultaneously: an error in any one of
them would move at least one of the four. In particular Cmax and Tmax pin the
absorption chain (`nn`, `mtt`, `ka`), the half-life pins the disposition
micro-constants, and AUC(0-12) pins `CL/F` together with the molar dose
conversion.

# Assumptions and deviations

* **Inter-occasion variability is implemented, with two occasions.** Bihorel
  2023 places variability on the mean transit time as IOV rather than IIV
  (Results, stage 3), and the Table 2 note states that occasions were "labeled
  as occasion 1 and occasion 2" with most participants having only one. The
  model therefore carries `etaiov_mtt_1` and `etaiov_mtt_2`, the second fixed
  equal to the first to encode the shared NONMEM `$OMEGA BLOCK(1) SAME` that a
  single reported magnitude implies. Simulations here use `OCC = 1`
  throughout, which is the correct setting for every participant with COVID-19
  and for the single-ascending-dose part of the phase I trial.
* **Food status in the COVID-19 cohort is a modelling construct, not data.**
  Food status was not collected in any phase II or III trial (Table 1). Bihorel
  2023 assigned it by mixture modelling and, after a sensitivity analysis,
  fixed at 25% the fraction of unknown-food-status participants treated as
  having eaten a high-fat meal; that assignment was then hard-coded into the
  analysis dataset. The stochastic cohort above draws
  `FED_HIGHFAT ~ Bernoulli(0.25)` to match. The paper's own Limitations note
  that this assignment "may be confounded by disease status or other unknown
  covariates" and "could have impacted the estimation of other covariate
  effects".
* **Body weight and BMI correlation is assumed.** Table 1 reports both margins
  but not their joint distribution. The virtual cohort draws them from a
  bivariate normal with rho = 0.8, truncated to the reported ranges. Only the
  weight margin affects AUC (CL/F depends on weight alone), so the assumption
  influences the Cmax spread rather than the exposure gates above.
* **Reduced variability structure for sparsely sampled participants is not
  reproduced.** Bihorel 2023 estimated IIV only on CL/F for participants
  contributing fewer than three NHC measurements, because there were
  insufficient data to support two IIV terms and one IOV term (Results, stages
  2 and 3). That is a per-record estimation device rather than a property of
  the final model, so the packaged model applies the full variability structure
  to every simulated subject. Simulated variability is therefore, if anything,
  slightly wider than the fitted model would produce for a sparse participant.
* **The hospitalisation effect on MTT is intentionally absent.** It was carried
  through stage 2 but was "close to 0 and poorly estimated (RSE: 87.2%)" once
  the transit-compartment absorption model was restored at stage 3, and it was
  the single relationship removed by the final backward elimination. No final
  estimate is published, so it cannot be encoded; see `covariatesDataExcluded`.
* **Screened-but-rejected covariates carry no coefficients.** Age, renal
  function, hepatic function, race and ethnicity were all evaluated and none
  was retained. The paper reports their effects only as forest-plot AUC(0-12)
  geometric mean ratios (Figure 3), which are geometric means over shrunken
  empirical Bayes estimates rather than typical-value contrasts and therefore
  cannot be inverted into coefficients. They are documented in
  `covariatesDataExcluded` and deliberately not gated on here.
* **No allometry on Q/F or Vp/F.** This is the model as published; Bihorel 2023
  Limitations flags the omission explicitly as the reason the model should not
  be extrapolated to children.
* **The analytical transit chain assumes one active dose.** `rxode2::transit()`
  evaluates the Savic input rate from the most recent dose only. With a mean
  transit time of 0.435 h fasted and 2.27 h after a high-fat meal, the chain is
  effectively exhausted long before the next dose 12 h later, so superposition
  error is negligible at the regimens simulated here. It would not be for a
  regimen whose dosing interval approached the transit time.
* **No parameter value came from anywhere other than the paper's own text and
  tables.** No figure digitisation, no author correspondence and no upstream
  model were needed; Table 2 reports the complete final parameter set. The
  supplement (Tables S1-S3, Figures S1-S10) holds sampling schedules, the
  modified Child-Pugh criteria, sample-exclusion accounting and diagnostic
  plots, none of which carry a model parameter. A Crossref check found no
  erratum or correction notice for doi:10.1002/psp4.13031.
