---
title: "Levetiracetam (Bilbao-Meseguer 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Levetiracetam (Bilbao-Meseguer 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r meta}
mod_fun <- readModelDb("BilbaoMeseguer_2021_levetiracetam")
mod <- rxode2::rxode2(mod_fun)
```

* Citation: `r mod$reference`
* Description: `r mod$description`
* Article (open access): <https://doi.org/10.3390/pharmaceutics13101690>

Bilbao-Meseguer et al. (2021) characterised intravenous levetiracetam in 27
critically ill adults whose measured urinary creatinine clearance (CrCl)
exceeded 50 mL/min, ten of whom had augmented renal clearance (ARC, CrCl above
130 mL/min). The final model is a two-compartment linear model in which
clearance is the sum of a CrCl-independent arm (3.5 L/h) and a power function
of CrCl, `(CrCl/120)^2.5` L/h. The paper used the model to evaluate
probability of target attainment (PTA) for steady-state troughs across
500-2000 mg every 12 h or every 8 h (Tables 4 and 5).

# Population

Twenty-seven ICU patients were enrolled prospectively at two Spanish hospitals
(Araba University Hospital, Vitoria-Gasteiz; Doce de Octubre Hospital, Madrid)
in 2019-2020 (Section 2.1, Table 1). Median age was 60 years (range 23-81),
median weight 80 kg (58-115), and 67% were male. Diagnoses were haemorrhagic
stroke (37%), trauma (30%) and other neurological conditions (33%); median
APACHE II was 18 (5-35). Urinary CrCl, measured from a urine collection as
urine creatinine x urine flow / plasma creatinine and **not** normalised to
body surface area, had median 117 mL/min (54-239). Patients received 500, 1000
or 1500 mg every 12 h as a 30-min IV infusion (18 of 27 on 500 mg) and
contributed 158 steady-state plasma samples (median 6 per patient).

```{r population}
str(mod_fun()$population)
```

# Source trace

Every `ini()` value carries an in-file comment pointing at its source; the
table collects them.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl_nonren` | log(3.5) L/h | Table 3, theta_nr (final model) |
| `e_crcl_cl_renal` | 2.5 | Table 3, theta_r (final model) |
| `lvc` | log(20.7) L | Table 3, V1 (final model) |
| `lq` | log(31.9) L/h | Table 3, Q (final model) |
| `lvp` | log(33.5) L | Table 3, V2 (final model) |
| `etalcl` | log(0.327^2 + 1) | Table 3, IIV_CL 32.7% |
| `etalvc` | log(0.561^2 + 1) | Table 3, IIV_V1 56.1% |
| `propSd` | 0.223 | Table 3, RE_proportional 22.3% |
| `cl <- (cl_nonren + (CRCL/120)^e_crcl_cl_renal) * exp(etalcl)` | -- | Section 3.3 final-model equation; Table 3 row header `CL = theta_nr + (CrCl/120)^theta_r` |
| `vc <- exp(lvc + etalvc)` | -- | Section 3.3 final-model equation `V1 = 20.7 x exp(eta2)` |
| two-compartment linear ODEs | -- | Section 3.3 ("two-compartment linear model ... CL, V1, V2 and Q") |
| `Cc ~ prop(propSd)` | -- | Section 3.3 ("Residual variability was proportionally modelled") |

## The CrCl term is a power, not a slope

Table 3 prints the covariate model as `CL = theta_nr + (CrCl/120)^theta_r`,
with theta_r set as a superscript, and the Section 3.3 equation typesets
`(CrCl/120)^2.5`. A text extraction that flattens the superscript reads it as
a linear slope, `3.5 + 2.5 x CrCl/120`. The Discussion settles which one was
fitted: "mean levetiracetam CL increased from 4.5 L/h to 9.2 L/h in patients
with CrCl from 120 to 240 mL/min", and Ong et al.'s clearance is compared with
"4.1 L/h for a mean CrCl of 100 mL/min". Only the power form gives these
numbers (the linear reading gives 6.0, 8.5 and 5.6 L/h).

```{r typical-cl}
typical_cl <- function(crcl) 3.5 + (crcl / 120)^2.5
cl_check <- data.frame(
  CRCL = c(100, 120, 240),
  paper = c(4.1, 4.5, 9.2),
  power_form = typical_cl(c(100, 120, 240)),
  linear_form = 3.5 + 2.5 * c(100, 120, 240) / 120
)
knitr::kable(
  cl_check |>
    dplyr::rename(
      "CrCl (mL/min)" = CRCL,
      "Paper Discussion CL (L/h)" = paper,
      "(CrCl/120)^2.5 (L/h)" = power_form,
      "2.5 x CrCl/120 (L/h)" = linear_form
    ),
  digits = 2
)
stopifnot(
  all(abs(cl_check$power_form - cl_check$paper) < 0.05),
  all(abs(cl_check$linear_form - cl_check$paper) > 0.5)
)
```

The packaged model is checked against the same closed form below by solving it
with all random effects at zero.

# Typical-value steady state and an exact clearance identity

At steady state the area under one dosing interval equals `Dose / CL`
exactly, whatever the distribution kinetics, so a typical-value solve over one
interval checks the clearance equation as implemented.

```{r ss-identity}
mod_typ <- rxode2::zeroRe(mod)
crcl_grid <- c(80, 120, 160, 200, 240)
ss_times <- sort(unique(c(seq(0, 0.5, by = 0.05), seq(0.5, 12, by = 0.1))))
ev_ss <- bind_rows(lapply(seq_along(crcl_grid), function(i) {
  bind_rows(
    data.frame(
      id = i, time = 0, amt = 1000, rate = 2000, ii = 12, ss = 1,
      evid = 1, cmt = "central"
    ),
    data.frame(
      id = i, time = ss_times, amt = 0, rate = 0, ii = 0, ss = 0,
      evid = 0, cmt = "central"
    )
  ) |>
    mutate(CRCL = crcl_grid[i])
}))
# zeroRe() leaves no omega; rxode2 warns about that for a multi-subject solve.
muffle_no_omega <- function(w) {
  if (grepl("omega", conditionMessage(w))) invokeRestart("muffleWarning")
}
sim_ss <- withCallingHandlers(
  rxode2::rxSolve(
    mod_typ, ev_ss,
    keep = "CRCL", returnType = "data.frame",
    rtol = 1e-10, atol = 1e-12, ssRtol = 1e-10, ssAtol = 1e-12,
    maxsteps = 1e6
  ),
  warning = muffle_no_omega
)

ss_dat <- sim_ss |> filter(!is.na(Cc)) |> mutate(dose = 1000)
conc_ss <- PKNCA::PKNCAconc(ss_dat, Cc ~ time | CRCL + id)
dose_ss <- PKNCA::PKNCAdose(
  ev_ss |> filter(evid == 1) |> mutate(dose = amt),
  dose ~ time | CRCL + id
)
res_ss <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  conc_ss, dose_ss,
  intervals = data.frame(start = 0, end = 12, auclast = TRUE, cmax = TRUE)
))
auc_ss <- as.data.frame(res_ss) |>
  filter(PPTESTCD == "auclast") |>
  select(CRCL, auc = PPORRES) |>
  left_join(
    sim_ss |> group_by(CRCL) |> summarise(cl = first(cl), .groups = "drop"),
    by = "CRCL"
  ) |>
  mutate(
    cl_closed_form = typical_cl(CRCL),
    dose_over_cl = 1000 / cl,
    rel_err = auc / dose_over_cl - 1
  )
knitr::kable(
  auc_ss |>
    dplyr::rename(
      "CrCl (mL/min)" = CRCL,
      "AUC0-12,ss (mg*h/L)" = auc,
      "Model CL (L/h)" = cl,
      "Closed-form CL (L/h)" = cl_closed_form,
      "Dose/CL (mg*h/L)" = dose_over_cl,
      "Relative error" = rel_err
    ),
  digits = 4
)
stopifnot(
  # The implemented clearance equals the Section 3.3 closed form.
  max(abs(auc_ss$cl / auc_ss$cl_closed_form - 1)) < 1e-8,
  # AUCtau = Dose/CL at steady state; the residual is the linear-up/log-down
  # trapezoid error on the 0.05-0.1 h grid (measured about 0.01%).
  max(abs(auc_ss$rel_err)) < 0.005
)
```

# Replicate published results

## Tables 4 and 5 -- probability of target attainment

The paper simulated 1000 virtual subjects at each of five CrCl values
(80-240 mL/min) and reported the percentage whose steady-state trough
exceeded 6, 12 and 46 mg/L, for 30-min infusions every 12 h (Table 4) and
every 8 h (Table 5), plus a 2-h infusion of 2000 mg every 8 h at 240 mL/min.

The replication below is deterministic. Instead of random draws it uses a
crossed grid of 14 x 14 standard-normal quantiles for the two random effects
(196 virtual subjects per CrCl value), solved with `zeroRe()` and the etas
supplied as data. The model is linear, so one 1000 mg solve per
regimen-CrCl pair is scaled to each dose.

```{r pta-sim}
pub_pta <- read.table(text = "
tau dur CRCL dose p6 p12 p46
12 0.5 80 500 62 12 0
12 0.5 80 1000 93 60 0
12 0.5 80 1500 99 85 3
12 0.5 80 2000 100 94 14
12 0.5 120 500 43 6 0
12 0.5 120 1000 86 43 0
12 0.5 120 1500 95 72 2
12 0.5 120 2000 98 85 6
12 0.5 160 500 22 1 0
12 0.5 160 1000 67 22 0
12 0.5 160 1500 87 51 0
12 0.5 160 2000 94 69 2
12 0.5 200 1000 39 6 0
12 0.5 200 1500 68 25 0
12 0.5 200 2000 80 42 0
12 0.5 240 1500 37 7 0
12 0.5 240 2000 55 15 0
8 0.5 80 500 94 51 0
8 0.5 80 1000 100 93 5
8 0.5 80 1500 100 99 31
8 0.5 120 500 84 33 0
8 0.5 120 1000 99 84 2
8 0.5 120 1500 100 96 17
8 0.5 160 500 65 12 0
8 0.5 160 1000 94 65 0
8 0.5 160 1500 99 89 5
8 0.5 160 2000 100 97 17
8 0.5 200 500 38 4 0
8 0.5 200 1000 83 39 0
8 0.5 200 1500 95 69 1
8 0.5 200 2000 98 84 5
8 0.5 240 1000 61 15 0
8 0.5 240 1500 80 38 0
8 0.5 240 2000 89 59 1
8 2 240 2000 94 67 1
", header = TRUE)

n_grid <- 14
z <- qnorm((seq_len(n_grid) - 0.5) / n_grid)
eta_grid <- expand.grid(zcl = z, zvc = z)
```

```{r pta-helpers}
om <- mod$omega
sd_cl <- sqrt(om["etalcl", "etalcl"])
sd_vc <- sqrt(om["etalvc", "etalvc"])

# Steady-state 1000 mg trough for every grid subject, regimen and CrCl.
# `crcl_in` lets the same model be driven with a transformed covariate.
solve_troughs <- function(crcl_map = identity) {
  regimens <- unique(pub_pta[, c("tau", "dur")])
  cohort <- tidyr::crossing(regimens, CRCL = crcl_grid, eta_grid) |>
    mutate(
      id = row_number(),
      etalcl = zcl * sd_cl,
      etalvc = zvc * sd_vc
    )
  ev <- bind_rows(
    cohort |> transmute(
      id, time = 0, amt = 1000, rate = 1000 / dur, ii = tau, ss = 1,
      evid = 1, cmt = "central", CRCL_model = crcl_map(CRCL), etalcl, etalvc
    ),
    cohort |> transmute(
      id, time = tau, amt = 0, rate = 0, ii = 0, ss = 0,
      evid = 0, cmt = "central", CRCL_model = crcl_map(CRCL), etalcl, etalvc
    )
  ) |>
    rename(CRCL = CRCL_model) |>
    arrange(id, time, desc(evid))
  sim <- withCallingHandlers(
    rxode2::rxSolve(mod_typ, ev, returnType = "data.frame", maxsteps = 1e6),
    warning = muffle_no_omega
  )
  sim |>
    select(id, Cc) |>
    left_join(cohort |> select(id, tau, dur, CRCL), by = "id")
}

# PTA (%) for each published row. `resid = TRUE` integrates the proportional
# residual error analytically; `resid = FALSE` uses the individual predictions.
score_pta <- function(troughs, resid) {
  pub_pta |>
    rowwise() |>
    mutate(
      cmin = list(troughs$Cc[troughs$tau == tau & troughs$dur == dur &
        troughs$CRCL == CRCL] * dose / 1000),
      s6 = if (resid) 100 * mean(pnorm(6, cmin, 0.223 * cmin, lower.tail = FALSE)) else 100 * mean(cmin > 6),
      s12 = if (resid) 100 * mean(pnorm(12, cmin, 0.223 * cmin, lower.tail = FALSE)) else 100 * mean(cmin > 12),
      s46 = if (resid) 100 * mean(pnorm(46, cmin, 0.223 * cmin, lower.tail = FALSE)) else 100 * mean(cmin > 46)
    ) |>
    ungroup() |>
    select(-cmin)
}

pta_diff <- function(tab) {
  with(tab, c(s6 - p6, s12 - p12, s46 - p46))
}
```

```{r pta-run}
troughs <- solve_troughs()
pta_ipred <- score_pta(troughs, resid = FALSE)
pta_resid <- score_pta(troughs, resid = TRUE)

# Alternative (rejected) reading of the covariate model: 3.5 + 2.5 * CrCl/120.
# (x/120)^2.5 == 2.5 * CrCl/120 when x = 120 * (2.5 * CrCl/120)^(1/2.5), so
# the packaged model can be driven on the linear reading without editing it.
troughs_linear <- solve_troughs(function(crcl) 120 * (2.5 * crcl / 120)^(1 / 2.5))
pta_linear <- score_pta(troughs_linear, resid = FALSE)

pta_summary <- data.frame(
  variant = c(
    "Power form, individual predictions",
    "Power form, with residual error",
    "Linear reading, individual predictions"
  ),
  d = I(list(pta_diff(pta_ipred), pta_diff(pta_resid), pta_diff(pta_linear)))
) |>
  rowwise() |>
  mutate(
    rmse = sqrt(mean(d^2)),
    median_abs = median(abs(d)),
    p90_abs = unname(quantile(abs(d), 0.9)),
    max_abs = max(abs(d))
  ) |>
  ungroup() |>
  select(-d)
knitr::kable(
  pta_summary |>
    dplyr::rename(
      "Variant" = variant,
      "RMSE (points)" = rmse,
      "Median |diff|" = median_abs,
      "90th pct |diff|" = p90_abs,
      "Max |diff|" = max_abs
    ),
  digits = 2
)
```

Over all `r nrow(pub_pta) * 3` published PTA cells the packaged model
reproduces Tables 4 and 5 with an RMSE of
`r round(pta_summary$rmse[1], 1)` percentage points using individual
predictions, and `r round(pta_summary$rmse[2], 1)` points when the residual
error is included. The paper's PTA therefore appears to have been computed
without residual error, but either way the agreement is close to the Monte
Carlo noise of a 1000-subject simulation (a binomial standard error of up to
1.6 points). The linear reading of the covariate term misses with an RMSE of
`r round(pta_summary$rmse[3], 1)` points.

```{r pta-gate}
d_ipred <- pta_diff(pta_ipred)
d_linear <- pta_diff(pta_linear)
stopifnot(
  # The eta grid is deterministic, so these are exact reruns, not draws.
  median(abs(d_ipred)) < 2.5,
  quantile(abs(d_ipred), 0.9) < 5,
  # Mutation control: the flattened-superscript reading must fail the same gate.
  sqrt(mean(d_linear^2)) > 3 * sqrt(mean(d_ipred^2))
)
```

```{r pta-table}
knitr::kable(
  pta_ipred |>
    mutate(
      regimen = sprintf("q%dh, %g-min infusion", tau, dur * 60),
      across(c(s6, s12, s46), \(x) round(x))
    ) |>
    select(regimen, CRCL, dose, p6, s6, p12, s12, p46, s46) |>
    dplyr::rename(
      "Regimen" = regimen,
      "CrCl (mL/min)" = CRCL,
      "Dose (mg)" = dose,
      ">6 mg/L paper" = p6,
      ">6 mg/L sim" = s6,
      ">12 mg/L paper" = p12,
      ">12 mg/L sim" = s12,
      ">46 mg/L paper" = p46,
      ">46 mg/L sim" = s46
    ),
  caption = "Replicates Tables 4 and 5 of Bilbao-Meseguer 2021 (PTA, %)."
)
```

```{r pta-figure}
pta_ipred |>
  select(tau, dur, CRCL, dose, p12, s12) |>
  filter(dur == 0.5) |>
  mutate(regimen = sprintf("every %d h", tau)) |>
  ggplot(aes(dose, s12, colour = factor(CRCL))) +
  geom_line() +
  geom_point(aes(y = p12), shape = 4, size = 2.5) +
  facet_wrap(~regimen) +
  geom_hline(yintercept = 80, linetype = "dashed") +
  labs(
    x = "Dose (mg, 30-min infusion)",
    y = "PTA, steady-state trough > 12 mg/L (%)",
    colour = "CrCl (mL/min)",
    title = "PTA for the 12 mg/L trough target",
    subtitle = "Lines: packaged model; crosses: Tables 4 and 5 of Bilbao-Meseguer 2021"
  ) +
  theme_minimal()
```

## Figure 4 -- clearance against CrCl

Figure 4 overlays the 5th, 50th and 95th percentiles of clearance from 1000
simulations at each CrCl value. Clearance is log-normal about the typical
value, so those percentiles follow in closed form from `etalcl`.

```{r figure-4}
fig4 <- data.frame(CRCL = seq(40, 240, by = 5)) |>
  mutate(
    p50 = typical_cl(CRCL),
    p05 = p50 * exp(qnorm(0.05) * sd_cl),
    p95 = p50 * exp(qnorm(0.95) * sd_cl)
  )
ggplot(fig4, aes(CRCL)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "grey80") +
  geom_line(aes(y = p50)) +
  labs(
    x = "CrCl (mL/min)",
    y = "Levetiracetam CL (L/h)",
    title = "Figure 4 -- clearance vs creatinine clearance",
    subtitle = "Replicates Bilbao-Meseguer 2021 Figure 4: median and 5th-95th percentile"
  ) +
  theme_minimal()
```

## Figures 1 and 5 -- steady-state profiles

Figure 1 shows the observed steady-state profiles and Figure 5 the
prediction-corrected VPC over one 12-h interval. The virtual cohort below
reproduces the study design: 500 mg every 12 h as a 30-min infusion, sampled
over the interval at steady state, with CrCl spread over the observed range.

```{r cohort}
set.seed(20210101)
n_arm <- 200
cohort <- bind_rows(
  data.frame(group = "No ARC", CRCL = runif(n_arm, 54, 130)),
  data.frame(group = "ARC", CRCL = runif(n_arm, 130, 239))
) |>
  mutate(id = row_number())

obs_times <- c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 5, 6, 7, 8, 10, 12)
ev_vpc <- bind_rows(
  cohort |> transmute(
    id, time = 0, amt = 500, rate = 1000, ii = 12, ss = 1,
    evid = 1, cmt = "central", CRCL
  ),
  tidyr::crossing(cohort, time = obs_times) |> transmute(
    id, time, amt = 0, rate = 0, ii = 0, ss = 0,
    evid = 0, cmt = "central", CRCL
  )
) |>
  arrange(id, time, desc(evid))
stopifnot(!anyDuplicated(ev_vpc[, c("id", "time", "evid")]))

rxode2::rxSetSeed(20210102)
sim_vpc <- rxode2::rxSolve(
  mod, ev_vpc,
  keep = "CRCL", returnType = "data.frame", maxsteps = 1e6
) |>
  left_join(cohort |> select(id, group), by = "id")
```

```{r figure-1-5}
vpc_q <- sim_vpc |>
  filter(!is.na(sim)) |>
  group_by(time) |>
  summarise(
    q10 = quantile(sim, 0.10),
    q50 = quantile(sim, 0.50),
    q90 = quantile(sim, 0.90),
    .groups = "drop"
  )
ggplot(vpc_q, aes(time, q50)) +
  geom_ribbon(aes(ymin = q10, ymax = q90), fill = "grey80") +
  geom_line() +
  geom_hline(yintercept = c(6, 12, 46), linetype = "dashed", colour = "grey40") +
  labs(
    x = "Time after dose (h)",
    y = "Levetiracetam (mg/L)",
    title = "Steady-state profile, 500 mg every 12 h",
    subtitle = "Median and 10th-90th percentile of simulated observations (cf. Figures 1 and 5)"
  ) +
  theme_minimal()
```

The simulated median is
`r round(vpc_q$q50[vpc_q$time == 0.5], 1)` mg/L at the end of the infusion and
`r round(vpc_q$q50[vpc_q$time == 12], 1)` mg/L at 12 h. The median line of the
Figure 5 pcVPC runs from about 23 mg/L at 0.5 h to about 6 mg/L at 12 h
(read from the figure by the maintainers). A prediction-corrected VPC is
normalised to the typical prediction of each observed record, so only the
shape and rough magnitude are comparable. The virtual cohort is also half ARC
(37% in the study), which pulls the simulated trough down.

# PKNCA against the published NCA (Table 2)

Table 2 reports the observed steady-state NCA, as group means, for patients
without and with ARC. The dose mix within each group is not reported, so the
comparison uses the dose-normalised quantities (`Cmax/D`, `AUC12/D`),
clearance and half-life, which do not depend on it.

```{r nca}
nca_dat <- sim_vpc |>
  filter(!is.na(Cc)) |>
  select(id, group, time, Cc)
conc_obj <- PKNCA::PKNCAconc(nca_dat, Cc ~ time | group + id)
dose_obj <- PKNCA::PKNCAdose(
  ev_vpc |>
    filter(evid == 1) |>
    left_join(cohort |> select(id, group), by = "id") |>
    mutate(dose = amt),
  dose ~ time | group + id
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  conc_obj, dose_obj,
  intervals = data.frame(
    start = 0, end = 12,
    cmax.dn = TRUE, auclast.dn = TRUE, cl.last = TRUE, half.life = TRUE
  )
))

sim_means <- as.data.frame(nca_res) |>
  filter(PPTESTCD %in% c("cmax.dn", "auclast.dn", "cl.last", "half.life")) |>
  group_by(group, PPTESTCD) |>
  summarise(PPORRES = mean(PPORRES, na.rm = TRUE), .groups = "drop")

ref_tab2 <- data.frame(
  group = c("No ARC", "ARC"),
  cmax.dn = c(0.053, 0.036),
  auclast.dn = c(0.267, 0.182),
  cl.last = c(4.28, 6.51),
  half.life = c(8.86, 7.25)
)

# ncaParamLabel() has no friendly label for the dose-normalised and
# clearance codes; it warns and returns the code, which is relabelled below.
cmp <- withCallingHandlers(
  nlmixr2lib::ncaComparisonTable(
    simulated = sim_means,
    reference = ref_tab2,
    by = "group",
    units = c(cmax.dn = "1/L", auclast.dn = "h/L", cl.last = "L/h", half.life = "h")
  ),
  warning = function(w) {
    if (grepl("unknown PKNCA code", conditionMessage(w))) invokeRestart("muffleWarning")
  }
)
cmp[["NCA parameter"]] <- dplyr::recode(
  cmp[["NCA parameter"]],
  "cmax.dn (1/L)" = "Cmax/D (1/L)",
  "auclast.dn (h/L)" = "AUC12/D (h/L)",
  "cl.last (L/h)" = "CL = D/AUC12 (L/h)"
)
knitr::kable(
  cmp |> dplyr::rename("Group" = group),
  caption = "Simulated means vs Table 2 of Bilbao-Meseguer 2021 (means)."
)
```

```{r nca-gate}
cl_sim <- sim_means |> filter(PPTESTCD == "cl.last")
cl_ref <- c("No ARC" = 4.28, "ARC" = 6.51)
cl_ratio <- cl_sim$PPORRES / cl_ref[cl_sim$group]
stopifnot(
  # Group-mean clearance of 200 subjects: a mis-transcribed arm or exponent
  # moves it by far more than 25%.
  all(abs(cl_ratio - 1) < 0.25),
  # ARC patients clear faster, the paper's central finding.
  cl_sim$PPORRES[cl_sim$group == "ARC"] > 1.2 * cl_sim$PPORRES[cl_sim$group == "No ARC"]
)
```

`r if (!is.null(attr(cmp, "footnote"))) attr(cmp, "footnote")`

Clearance and dose-normalised AUC agree with Table 2 within the uncertainty of
a 17- and a 10-patient group mean (the Table 2 standard deviations are 30-50%
of the means). Half-life is the least comparable quantity and is the row most likely to be
flagged: the paper's Phoenix estimates come from a 12-h interval of a
two-compartment drug and carry standard deviations of 57-69% of the mean, and
the simulated `half.life` is fitted to the tail of the same short interval.
The simulated no-ARC half-life also depends on how the virtual cohort spreads
CrCl within 54-130 mL/min, which the paper does not report.

# Assumptions and deviations

* **Covariate form.** The final CL model is taken as
  `CL = (3.5 + (CrCl/120)^2.5) x exp(eta)`, the typeset form of the
  Section 3.3 equation and the Table 3 header. The alternative linear reading
  `3.5 + 2.5 x CrCl/120` contradicts the Discussion's worked clearances and
  misses Tables 4-5 by much more (above). The renal arm therefore has an
  implicit coefficient of 1 L/h at 120 mL/min, and the 120 mL/min divisor is
  the value printed in the equation rather than the Table 1 median (117).
* **CrCl units.** `CRCL` for this model is the measured urinary creatinine
  clearance in raw mL/min, not normalised to 1.73 m^2. Supplying a
  BSA-normalised or Cockcroft-Gault value changes the prediction; with an
  exponent of 2.5 the renal arm is very sensitive to the covariate scale.
* **Extrapolation.** The fitted range of CrCl is 54-239 mL/min. The power
  term grows steeply, so predictions outside that range (in particular
  CrCl below 50 mL/min, which was an exclusion criterion) are extrapolations.
* **IIV scale.** The Table 3 IIV percentages are converted with
  `omega^2 = log(CV^2 + 1)`. Reading them as `sqrt(omega)` instead
  (omega^2 = 0.107 and 0.315) changes the PTA replication RMSE by less than
  0.2 points, so Tables 4 and 5 cannot distinguish the two readings.
* **Residual error in the PTA.** The paper does not say whether its PTA
  included residual error. Individual predictions reproduce Tables 4-5
  slightly better (RMSE 1.5 vs 2.4 points).
* **Total volume in the Discussion.** The Discussion quotes a total volume of
  distribution of 54.9 L, which is the sum of the base-model V1 and V2
  (20.8 + 34.1). The final-model sum is 54.2 L (20.7 + 33.5); the packaged
  model uses the final-model values.
* **Table 2 narrative.** Section 3.2 states that dose-normalised Cmax was
  significantly *higher* in ARC patients, while Table 2 shows it lower
  (0.036 vs 0.053 L^-1, starred as significant). The table values are used.
* **Virtual cohort for Table 2.** CrCl was drawn uniformly within 54-130 and
  130-239 mL/min for the two groups because the paper does not report the
  within-group distributions; all subjects received 500 mg every 12 h.
* **Screened covariates.** Trauma diagnosis and APACHE II were significant on
  V1 in forward inclusion but were removed at backward elimination, and the
  categorical ARC flag was replaced by continuous CrCl. They are recorded in
  the model's `covariatesDataExcluded` metadata and do not enter the model.
