---
title: "Buprenorphine (Bjornsson 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Buprenorphine (Bjornsson 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Bjornsson_2023_buprenorphine"))$reference`
* Article: <https://doi.org/10.1007/s40262-023-01288-6>
* Trial registrations: ISRCTN41550730, ISRCTN24987553, NCT02611752, NCT02710526

Buprenorphine (BPN) is a partial mu-opioid receptor agonist used to treat opioid
use disorder (OUD). CAM2038 is a subcutaneous extended-release BPN depot built on
FluidCrystal injection technology: a lipid solution that transforms in situ into a
liquid-crystalline gel and releases BPN as the depot matrix biodegrades. It is
supplied in a weekly (Q1W, 8-32 mg) and a monthly (Q4W, 64-160 mg) formulation.

Bjornsson 2023 pooled four trials into a **single population PK model covering
four routes of administration** - intravenous BPN, sublingual (SL) BPN tablets,
and subcutaneous CAM2038 Q1W and Q4W. Because the four routes were fitted jointly
and share one disposition model, they are packaged here as **one** model file
(per the replicate-the-author's-structure policy) with six absorption depots.

```{r description}
strwrap(rxode2::rxode2(readModelDb("Bjornsson_2023_buprenorphine"))$description,
        width = 78) |> cat(sep = "\n")
```

# Population

The analysis pooled two Phase 1 trials in healthy participants (Trials 1 and 2)
and two Phase 2 trials in participants with OUD (Trials 3 and 4). The final model
included **252 individuals and 10,658 BPN plasma concentration observations**
(Results 3.1.1) - the union of the initial-model dataset (236 participants,
10,260 observations) and 16 additional participants (398 observations) who
received 160 mg CAM2038 Q4W and became available after the initial model was
built. That update re-estimated only the CAM2038 Q4W absorption parameters.

Baseline characteristics (Table 1, `All, N = 236` column): median age 35 years
(range 18-65), median body weight 72.2 kg (50.4-127), median BMI 24.4 kg/m^2
(16.8-34.9), 39 % female, 66 % White / 22 % Black or African American / 6 % Asian
/ 2 % Mixed / 5 % Other, median creatinine clearance 112 mL/min (65.1-209).
Approximately 85 % of participants had normal renal function and 15 % mild
impairment; none had moderate or severe impairment. 62 % were healthy and 38 %
had OUD. Table 1 publishes no combined `N = 252` demographic column, so the
`population` metadata records the union of the two published columns.

```{r population}
pop <- readModelDb("Bjornsson_2023_buprenorphine")()$population
str(pop[c("species", "n_subjects", "n_studies", "age_range", "weight_range",
          "sex_female_pct", "disease_state")], width = 78, strict.width = "wrap")
```

# Model structure

Disposition is three-compartment with first-order elimination from `central`
(Figure 1). Every extravascular route absorbs through **two parallel depot
pathways**, one fast and one slow, with the administered dose split between them
by an estimated logit-transformed fraction:

| Route | Fast pathway | Slow pathway | Dose fraction to fast |
|---|---|---|---|
| IV | dose into `central`, F = 1 | -- | -- |
| SL BPN | `depot_sl1`: zero-order over `D_SL1` = 0.419 h after a 0.171 h lag, then `ka_SL1` = 1.72 1/h | `depot_sl2`: `ka_SL2` = 0.0875 1/h | `F_SL1` = 75.9 % of the bioavailable dose |
| CAM2038 Q1W | `depot_q1w1`: zero-order over `D_q1w1` = 10.1 h, then `ka_q1w1` = 0.0401 1/h | `depot_q1w2`: `ka_q1w2` = 0.00565 1/h | `Fq1w1` = 45.5 % |
| CAM2038 Q4W | `depot_q4w1`: `ka_q4w1` = 0.0441 1/h | `depot_q4w2`: `ka_q4w2` = 0.00166 1/h | `Fq4w1` = 8.63 % |

Bioavailability is 1 for IV and was **fixed to 1** for both CAM2038 formulations
(estimated close to the upper boundary; Results 3.1.1). SL bioavailability is
dose-dependent. To dose an extravascular route, supply the **same amount to both
depots of the pair** - the `f()` statements split it, exactly as the source
NONMEM model does with per-compartment F.

## Derived quantities reproduced from the parameter set

Several numbers the paper quotes in prose are pure consequences of the `ini()`
values, so they are an immediate arithmetic check on the transcription - no
simulation required.

```{r derived-checks}
mod  <- readModelDb("Bjornsson_2023_buprenorphine")
th   <- rxode2::rxode2(mod)$theta

hl <- function(nm) log(2) / exp(th[[nm]])
fsl <- function(dose) {
  p <- 1 / (1 + exp(-th[["logitfdepot_sl"]]))
  p * (dose / 16)^th[["e_dose_fdepot_sl"]]
}
cl <- function(age, wt) {
  exp(th[["lcl"]]) * (age / 35)^th[["e_age_cl"]] * (wt / 72.4)^th[["e_wt_cl"]]
}
expit <- function(x) 1 / (1 + exp(-x))
fq1w1 <- function(oud = 0, sexf = 0) {
  expit(th[["logitfdepot_q1w1"]] + th[["e_dis_oud_fdepot_q1w1"]] * oud +
          th[["e_sexf_fdepot_q1w1"]] * sexf)
}

tibble::tribble(
  ~Quantity,                                        ~Model,                       ~Paper,
  "Absorption half-life, ka_SL2 (h)",               hl("lka_sl2"),                7.9,
  "Absorption half-life, ka_q1w2 (h)",              hl("lka_q1w2"),               123,
  "Absorption half-life, ka_q4w2 (h)",              hl("lka_q4w2"),               418,
  "F_SL at 8 mg (%)",                               100 * fsl(8),                 18.1,
  "F_SL at 16 mg (%)",                              100 * fsl(16),                14.0,
  "F_SL at 24 mg (%)",                              100 * fsl(24),                12.0,
  "CL at age 18 y, 72.4 kg (L/h)",                  cl(18, 72.4),                 60.8,
  "CL at age 65 y, 72.4 kg (L/h)",                  cl(65, 72.4),                 45.1,
  "CL at 50 kg, age 35 y (L/h)",                    cl(35, 50),                   44.7,
  "CL at 100 kg, age 35 y (L/h)",                   cl(35, 100),                  59.5,
  "Vc, participants with OUD (L)",                  exp(th[["lvc"]]) * (1 + th[["e_dis_oud_vc"]]), 237,
  "Fq1w1, healthy male (%)",                        100 * fq1w1(0, 0),            45.5,
  "Fq1w1, male with OUD (%)",                       100 * fq1w1(1, 0),            29.9,
  "Fq1w1, healthy female (%)",                      100 * fq1w1(0, 1),            59.8
) |>
  mutate(`% diff` = 100 * (Model - Paper) / Paper) |>
  knitr::kable(digits = c(0, 3, 3, 2),
               caption = paste("Quantities quoted in Bjornsson 2023 prose,",
                               "recomputed from the packaged ini() values."))
```

The paper also states that "regardless of dose, 78.9 % of the CAM2038 Q1W dose
was absorbed after 7 days and 70.1 % of the CAM2038 Q4W dose was absorbed after
28 days" (Results 3.2, Figure S8). Both follow from the pathway split and the
slow-pathway rate constants.

```{r absorbed-fraction}
absorbed <- function(f_fast, ka_fast, ka_slow, t) {
  f_fast * (1 - exp(-ka_fast * t)) + (1 - f_fast) * (1 - exp(-ka_slow * t))
}
tibble::tibble(
  Formulation = c("CAM2038 Q1W at 7 days", "CAM2038 Q4W at 28 days"),
  Model = 100 * c(
    absorbed(expit(th[["logitfdepot_q1w1"]]), exp(th[["lka_q1w1"]]),
             exp(th[["lka_q1w2"]]), 7 * 24),
    absorbed(expit(th[["logitfdepot_q4w1"]]), exp(th[["lka_q4w1"]]),
             exp(th[["lka_q4w2"]]), 28 * 24)
  ),
  Paper = c(78.9, 70.1)
) |>
  knitr::kable(digits = 2, caption = "Percent of the depot dose absorbed (Figure S8).")
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Bjornsson_2023_buprenorphine.R` carries an in-file
comment naming its origin. The table collects them for review. All structural,
covariate, IIV, and residual-error values come from **Table 3** ("Parameter
estimates for the final BPN population PK model"); the equations come from the
Methods and Results narrative.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` | 52.1 L/h | Table 3, `CL` (RSE 1.80 %) |
| `lvc` | 64.3 L | Table 3, `Vc` (RSE 10.3 %); healthy reference |
| `lq` | 186 L/h | Table 3, `Q2` |
| `lvp` | 130 L | Table 3, `V2` |
| `lq2` | 60.3 L/h | Table 3, `Q3` |
| `lvp2` | 1580 L | Table 3, `V3` |
| `logitfdepot_sl` | 14.0 % | Table 3, `FSL` |
| `logitfdepot_sl1` | 75.9 % | Table 3, `FSL1` |
| `ltlag_sl1` | 0.171 h | Table 3, `tlag,SL1` |
| `ld1_sl1` | 0.419 h | Table 3, `DSL1` |
| `lka_sl1` | 1.72 1/h | Table 3, `ka,SL1` |
| `lka_sl2` | 0.0875 1/h | Table 3, `ka,SL2` |
| `logitfdepot_q1w1` | 45.5 % | Table 3, `Fq1w1` |
| `ld1_q1w1` | 10.1 h | Table 3, `Dq1w1` |
| `lka_q1w1` | 0.0401 1/h | Table 3, `ka,q1w1` |
| `lka_q1w2` | 0.00565 1/h | Table 3, `ka,q1w2` |
| `logitfdepot_q4w1` | 8.63 % | Table 3, `Fq4w1` |
| `lka_q4w1` | 0.0441 1/h | Table 3, `ka,q4w1` |
| `lka_q4w2` | 0.00166 1/h | Table 3, `ka,q4w2` |
| `e_age_cl` | -0.233 | Table 3, "Age covariate on CL" |
| `e_wt_cl` | 0.413 | Table 3, "WT covariate on CL" |
| `e_dis_oud_vc` | 2.69 | Table 3, "Population covariate on Vc", footnote a |
| `e_dis_oud_fdepot_q1w1` | -0.671 | Table 3, "Population covariate on Fq1w1", footnote a |
| `e_sexf_fdepot_q1w1` | 0.576 | Table 3, "Sex covariate on Fq1w1", footnote b |
| `e_dose_fdepot_sl` | -0.371 | Table 3, "Dose dependency on FSL" |
| All 11 `eta` variances | Table 3 `IIV / CV (%)` column | Table 3 (see Assumptions on how the column is read) |
| `propSd` | 27.7 % | Table 3, `sigma_prop` |
| Three-compartment disposition, first-order elimination | n/a | Results 3.1.1; Figure 1 |
| Dual parallel absorption per route; zero-order `D` and lag on the fast SL / Q1W pathways | n/a | Results 3.1.1; Figure 1 |
| `F_SL = 0.14 * (Dose / 16)^-0.371` | n/a | Results 3.1.3 (printed equation) |
| `CL = 52.1 * (Age/35)^-0.233 * (WT/72.4)^0.413` | n/a | Results 3.1.3 (printed equation) |
| `Vc = 64.3 * (1 + 2.69 * DIS_OUD)` | n/a | Methods Eq. 5 (categorical fractional difference) + Results 3.1.3 (64.3 vs 237 L) |
| Additive shifts on `logit(Fq1w1)` for OUD and female sex | n/a | Methods Eqs. 2-4 + Results 3.1.3 direction statements (see Assumptions) |
| `Fq1w1 = 0` for thigh injections | n/a | Results 3.1.1 ("close to 0 and subsequently fixed to 0") |
| CAM2038 F fixed to 1 | n/a | Results 3.1.1 |
| Proportional residual error (additive on log-transformed data) | n/a | Results 3.1.1 |
| Reference age 35 y, weight 72.4 kg, SL dose 16 mg | n/a | Methods 2.4 (typical individual); Results 3.1.3 |

# Virtual cohorts

Original observed data are not publicly available. Two cohorts are used below.

The **replication cohort** matches the paper's own simulation design: Methods 2.4
states the simulations were "based on predictions for a typical healthy
individual (male participant, aged 35 years, 72.4 kg body weight), and were
replicated 1000 times". Covariates are therefore held at those typical values and
only the inter-individual random effects vary. Cohort size is 200 per arm (the
nlmixr2lib cap); the paper used 1000.

The **demographic cohort** spreads covariates over the Table 1 ranges and is used
only to illustrate the retained covariate effects.

```{r cohort-helpers}
set.seed(20230816)
N_ARM <- 200L

# Depot pairs per route, and which member carries a modelled zero-order duration
# (rate = -2 tells rxode2 to use dur(); the SL/Q1W fast pathways need it).
route_depots <- list(
  sl  = list(cmt = c("depot_sl1",  "depot_sl2"),  rate = c(-2, 0)),
  q1w = list(cmt = c("depot_q1w1", "depot_q1w2"), rate = c(-2, 0)),
  q4w = list(cmt = c("depot_q4w1", "depot_q4w2"), rate = c( 0, 0))
)

# One arm: `ndose` doses of `dose` mg every `tau` h, observed on `obs_time`.
make_arm <- function(route, dose, tau, ndose, obs_time, label,
                     id_offset = 0L, n = N_ARM,
                     age = 35, wt = 72.4, sexf = 0, oud = 0, thigh = 0) {
  dp <- route_depots[[route]]
  # Computed outside mutate(): inside, `route` would resolve to the new column.
  dose_sl <- if (identical(route, "sl")) dose else 16
  dos <- tibble::tibble(time = 0, amt = dose, evid = 1L, cmt = dp$cmt,
                        rate = dp$rate, ii = tau, addl = ndose - 1L)
  obs <- tibble::tibble(time = obs_time, amt = NA_real_, evid = 0L,
                        cmt = "central", rate = 0, ii = 0, addl = 0L)
  tidyr::crossing(id = id_offset + seq_len(n), dplyr::bind_rows(dos, obs)) |>
    dplyr::mutate(
      treatment = label, route = route, dose_mg = dose, tau = tau,
      AGE = age, WT = wt, SEXF = sexf, DIS_OUD = oud, INJSITE_THIGH = thigh,
      DOSE_BPN_SL_MG = dose_sl
    ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}
```

## Replication cohort for Table 4 (steady state)

Table 4 reports simulated steady-state `Cmax,ss`, `Cav,ss`, `Ctrough,ss`, and the
`Cmax,ss / Ctrough,ss` ratio for 11 regimens. Dose counts below were chosen so
each regimen is fully converged to steady state: with these parameters the
typical-value `Cav,ss` changes by < 0.1 % between 20 and 40 daily SL doses,
between 8 and 26 weekly doses, and between 8 and 14 monthly doses.

```{r ss-events}
ss_regimens <- tibble::tribble(
  ~label,          ~route, ~dose, ~tau, ~ndose, ~step,
  "SL 8 mg",       "sl",      8,    24,     20,  0.25,
  "SL 16 mg",      "sl",     16,    24,     20,  0.25,
  "SL 24 mg",      "sl",     24,    24,     20,  0.25,
  "Q1W 8 mg",      "q1w",     8,   168,     12,  1,
  "Q1W 16 mg",     "q1w",    16,   168,     12,  1,
  "Q1W 24 mg",     "q1w",    24,   168,     12,  1,
  "Q1W 32 mg",     "q1w",    32,   168,     12,  1,
  "Q4W 64 mg",     "q4w",    64,   672,     10,  2,
  "Q4W 96 mg",     "q4w",    96,   672,     10,  2,
  "Q4W 128 mg",    "q4w",   128,   672,     10,  2,
  "Q4W 160 mg",    "q4w",   160,   672,     10,  2
)

ss_events <- dplyr::bind_rows(lapply(seq_len(nrow(ss_regimens)), function(i) {
  r <- ss_regimens[i, ]
  t_last <- r$tau * (r$ndose - 1L)
  make_arm(r$route, r$dose, r$tau, r$ndose,
           obs_time = seq(t_last, t_last + r$tau, by = r$step),
           label = r$label, id_offset = (i - 1L) * N_ARM)
}))
stopifnot(!anyDuplicated(unique(ss_events[, c("id", "time", "evid", "cmt")])))
ss_events <- ss_events |> mutate(treatment = factor(treatment, ss_regimens$label))
```

# Simulation

```{r simulate-ss}
sim_ss <- rxode2::rxSolve(
  mod, events = ss_events,
  keep = c("treatment", "route", "dose_mg", "tau")
) |> as.data.frame()
nrow(sim_ss)
```

# Replicate published figures

## Figure 3 - treatment initiation with CAM2038 Q1W

Figure 3 shows median typical-value profiles over the first treatment week for
16 mg, 16 + 8 mg (extra dose on day 3), 16 + 8 + 8 mg (extra doses on days 3 and
5), 24 mg, and 32 mg CAM2038 Q1W in individuals not previously exposed to BPN.
These are typical-value predictions, so the random effects are zeroed.

```{r figure-3}
mod_typ <- mod |> rxode2::zeroRe()

# Titration arms need per-dose amounts, so build them explicitly rather than
# with make_arm()'s single ii/addl block.
init_arm <- function(dose_times, doses, label) {
  dos <- tibble::tibble(
    time = rep(dose_times, each = 2), amt = rep(doses, each = 2), evid = 1L,
    cmt = rep(route_depots$q1w$cmt, times = length(doses)),
    rate = rep(route_depots$q1w$rate, times = length(doses))
  )
  obs <- tibble::tibble(time = seq(0, 168, by = 0.5), amt = NA_real_,
                        evid = 0L, cmt = "central", rate = 0)
  dplyr::bind_rows(dos, obs) |>
    dplyr::mutate(id = 1L, treatment = label, AGE = 35, WT = 72.4, SEXF = 0,
                  DIS_OUD = 0, INJSITE_THIGH = 0, DOSE_BPN_SL_MG = 16) |>
    dplyr::arrange(time, dplyr::desc(evid))
}

init_labels <- c("16 mg", "16 + 8 mg (d3)", "16 + 8 + 8 mg (d3, d5)",
                 "24 mg", "32 mg")
init_events <- dplyr::bind_rows(
  init_arm(0,               16,           init_labels[1]) |> mutate(id = 1L),
  init_arm(c(0, 72),        c(16, 8),     init_labels[2]) |> mutate(id = 2L),
  init_arm(c(0, 72, 120),   c(16, 8, 8),  init_labels[3]) |> mutate(id = 3L),
  init_arm(0,               24,           init_labels[4]) |> mutate(id = 4L),
  init_arm(0,               32,           init_labels[5]) |> mutate(id = 5L)
)

sim_init <- rxode2::rxSolve(mod_typ, events = init_events,
                            keep = "treatment") |> as.data.frame()

sim_init |>
  filter(!is.na(Cc)) |>
  mutate(treatment = factor(treatment, init_labels)) |>
  ggplot(aes(time / 24, Cc, colour = treatment)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 1.2, linetype = "dashed", colour = "grey40") +
  scale_x_continuous(breaks = 0:7) +
  labs(x = "Time (days)", y = "BPN plasma concentration (ng/mL)",
       colour = "CAM2038 Q1W",
       title = "Figure 3 - treatment initiation with CAM2038 Q1W",
       caption = paste("Replicates Figure 3 of Bjornsson 2023. Dashed line:",
                       "1.2 ng/mL opioid-blockade concentration.")) +
  theme(legend.position = "bottom")
```

Bjornsson 2023 Results 3.2 states that "BPN concentrations of 1.2 ng/mL were
reached approximately 5 h after the recommended starting dose of 16 mg CAM2038
Q1W". The remaining reference values below were read off the published Figure 3
by eye (its axis gridlines resolve to roughly 0.05 ng/mL), so they carry
digitisation uncertainty and are a shape check rather than an exact target.

```{r figure-3-checks}
# Values digitised from Figure 3 of Bjornsson 2023 (see caveat above). For the
# titration arms the figure's visually highest late peak is the one following
# the FINAL 8 mg dose, so `peak_from` gives the time of that dose.
fig3_ref <- tibble::tribble(
  ~treatment,                ~peak_from, ~ref_peak, ~ref_day7,
  "16 mg",                            0,      2.63,      0.52,
  "16 + 8 mg (d3)",                  72,      2.50,      1.10,
  "16 + 8 + 8 mg (d3, d5)",         120,      2.82,      1.95,
  "24 mg",                            0,      3.85,      0.79,
  "32 mg",                            0,      5.23,      1.04
)

sim_init |>
  filter(!is.na(Cc)) |>
  left_join(fig3_ref, by = "treatment") |>
  group_by(treatment, peak_from, ref_peak, ref_day7) |>
  summarise(
    last_peak = max(Cc[time >= peak_from[1]]),
    tmax_days = time[which.max(Cc)] / 24,
    `Time to 1.2 ng/mL (h)` = suppressWarnings(min(time[Cc >= 1.2])),
    day7 = Cc[which.min(abs(time - 168))],
    .groups = "drop"
  ) |>
  mutate(treatment = factor(treatment, init_labels)) |>
  arrange(treatment) |>
  select(treatment, last_peak, ref_peak, day7, ref_day7,
         `Time to 1.2 ng/mL (h)`, tmax_days) |>
  rename("CAM2038 Q1W regimen" = treatment,
         "Simulated final peak (ng/mL)" = last_peak,
         "Figure 3 final peak (ng/mL)" = ref_peak,
         "Simulated day 7 (ng/mL)" = day7,
         "Figure 3 day 7 (ng/mL)" = ref_day7,
         "Day-1 Tmax (days)" = tmax_days) |>
  knitr::kable(digits = 3,
               caption = paste("Typical-value initiation metrics versus values",
                               "digitised from Figure 3 of Bjornsson 2023."))
```

## Figure 5 - bridging from SL BPN to CAM2038 at steady state

Figure 5 overlays steady-state profiles for the established dose conversions:
16 mg daily SL BPN with 24 mg CAM2038 Q1W (panel a), 16 mg SL with 96 mg CAM2038
Q4W (panel b), and 24 mg Q1W with 96 mg Q4W (panel c). Time is shown relative to
the start of the last dosing interval, on the panel's own interval length.

```{r figure-5}
bridge <- tibble::tribble(
  ~panel,                                  ~treatment,
  "a: 16 mg SL vs 24 mg Q1W",              "SL 16 mg",
  "a: 16 mg SL vs 24 mg Q1W",              "Q1W 24 mg",
  "b: 16 mg SL vs 96 mg Q4W",              "SL 16 mg",
  "b: 16 mg SL vs 96 mg Q4W",              "Q4W 96 mg",
  "c: 24 mg Q1W vs 96 mg Q4W",             "Q1W 24 mg",
  "c: 24 mg Q1W vs 96 mg Q4W",             "Q4W 96 mg"
) |> mutate(treatment = factor(treatment, ss_regimens$label))

sim_ss |>
  filter(!is.na(Cc)) |>
  group_by(treatment, time) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
            .groups = "drop") |>
  group_by(treatment) |>
  mutate(t_rel = (time - min(time)) / 24) |>
  ungroup() |>
  inner_join(bridge, by = "treatment", relationship = "many-to-many") |>
  ggplot(aes(t_rel, Q50, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.18, colour = NA) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~panel, scales = "free_x") +
  labs(x = "Time within the steady-state dosing interval (days)",
       y = "BPN plasma concentration (ng/mL)", colour = NULL, fill = NULL,
       title = "Figure 5 - steady-state bridging between SL BPN and CAM2038",
       caption = paste("Replicates Figure 5 of Bjornsson 2023. Line = median,",
                       "band = 5th-95th percentile of 200 simulated subjects.")) +
  theme(legend.position = "bottom")
```

## Figure 7 and Table 5 - tapering after interrupting treatment

Figure 7 shows the decline after five doses of CAM2038, and Table 5 reports the
geometric-mean time for BPN to fall below 0.1 ng/mL for each Q1W and Q4W dose.

```{r taper-sim}
taper_regimens <- tibble::tribble(
  ~label,       ~route, ~dose, ~tau, ~ndose, ~follow_days,
  "16 mg Q1W",  "q1w",    16,   168,      5,   90,
  "24 mg Q1W",  "q1w",    24,   168,      5,   90,
  "32 mg Q1W",  "q1w",    32,   168,      5,   90,
  "64 mg Q4W",  "q4w",    64,   672,      5,  220,
  "96 mg Q4W",  "q4w",    96,   672,      5,  220,
  "128 mg Q4W", "q4w",   128,   672,      5,  220,
  "160 mg Q4W", "q4w",   160,   672,      5,  220
)

taper_events <- dplyr::bind_rows(lapply(seq_len(nrow(taper_regimens)), function(i) {
  r <- taper_regimens[i, ]
  t_last <- r$tau * (r$ndose - 1L)
  make_arm(r$route, r$dose, r$tau, r$ndose,
           obs_time = t_last + seq(0, r$follow_days * 24, by = 12),
           label = r$label, id_offset = (i - 1L) * N_ARM)
}))
stopifnot(!anyDuplicated(unique(taper_events[, c("id", "time", "evid", "cmt")])))

sim_taper <- rxode2::rxSolve(mod, events = taper_events,
                             keep = c("treatment", "route")) |> as.data.frame()
```

```{r figure-7}
sim_taper |>
  filter(!is.na(Cc), treatment %in% c("32 mg Q1W", "128 mg Q4W")) |>
  group_by(treatment, time) |>
  summarise(Q025 = quantile(Cc, 0.025), Q50 = median(Cc),
            Q975 = quantile(Cc, 0.975), .groups = "drop") |>
  group_by(treatment) |>
  mutate(days_after_last = (time - min(time)) / 24) |>
  ungroup() |>
  ggplot(aes(days_after_last, Q50)) +
  geom_ribbon(aes(ymin = Q025, ymax = Q975), alpha = 0.2) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 0.1, linetype = "dashed", colour = "firebrick") +
  facet_wrap(~treatment, scales = "free_x") +
  scale_y_log10() +
  labs(x = "Days after the last dose", y = "BPN plasma concentration (ng/mL)",
       title = "Figure 7 - decline after five doses of CAM2038",
       caption = paste("Replicates Figure 7 of Bjornsson 2023. Dashed line:",
                       "the 0.1 ng/mL threshold used in Table 5."))
```

```{r table-5}
gm <- function(x) exp(mean(log(x)))

taper_pub <- tibble::tribble(
  ~treatment,   ~pub_gm, ~pub_lo, ~pub_hi,
  "16 mg Q1W",     21.8,    10.9,    52.3,
  "24 mg Q1W",     25.6,    12.0,    60.9,
  "32 mg Q1W",     27.3,    13.0,    63.1,
  "64 mg Q4W",     82.3,    43.9,     150,
  "96 mg Q4W",     92.9,    50.4,     165,
  "128 mg Q4W",     101,    55.4,     183,
  "160 mg Q4W",     104,    56.9,     192
)

# Last time at or above 0.1 ng/mL; the next grid point is the first sustained
# time below the threshold. Concentrations decline monotonically over the
# follow-up window, so this is the crossing time to grid resolution (12 h).
first_below <- function(time, Cc, thresh = 0.1) {
  o <- order(time)
  time <- time[o]; Cc <- Cc[o]
  if (all(Cc < thresh)) return(0)
  if (Cc[length(Cc)] >= thresh) return(NA_real_)
  (time[max(which(Cc >= thresh)) + 1L] - time[1]) / 24
}

sim_taper |>
  filter(!is.na(Cc)) |>
  group_by(treatment, id) |>
  summarise(t_below = first_below(time, Cc), .groups = "drop") |>
  group_by(treatment) |>
  summarise(n_reached = sum(!is.na(t_below)),
            sim_gm = gm(t_below[!is.na(t_below)]),
            sim_lo = quantile(t_below, 0.025, na.rm = TRUE),
            sim_hi = quantile(t_below, 0.975, na.rm = TRUE), .groups = "drop") |>
  left_join(taper_pub, by = "treatment") |>
  mutate(`% diff` = 100 * (sim_gm - pub_gm) / pub_gm,
         treatment = factor(treatment, taper_regimens$label)) |>
  arrange(treatment) |>
  select(treatment, sim_gm, pub_gm, `% diff`, sim_lo, sim_hi, pub_lo, pub_hi) |>
  rename("Regimen" = treatment,
         "Simulated geo. mean (d)" = sim_gm, "Paper geo. mean (d)" = pub_gm,
         "Simulated 2.5th (d)" = sim_lo, "Simulated 97.5th (d)" = sim_hi,
         "Paper 95% PI low (d)" = pub_lo, "Paper 95% PI high (d)" = pub_hi) |>
  knitr::kable(digits = 1,
               caption = paste("Table 5 of Bjornsson 2023 - days for BPN to fall",
                               "below 0.1 ng/mL after five CAM2038 doses."))
```

# PKNCA validation

Steady-state NCA is computed on the final dosing interval of each regimen, with
the treatment label as the grouping variable so results can be compared with
Table 4 per regimen.

The paper's `Ctrough,ss` is taken from PKNCA's `cmin` over the steady-state
interval rather than `ctrough`: these profiles decline monotonically from their
peak to the end of the interval, so the interval minimum *is* the trough (for the
Q1W 16 mg arm the two agree to four significant figures), and PKNCA's `ctrough`
returns `NA` in this configuration. The `Cmax,ss / Ctrough,ss` ratio is therefore
formed per subject as `cmax / cmin` and then aggregated, rather than taken from
PKNCA's `ptr`.

```{r pknca}
sim_nca <- sim_ss |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

dose_df <- ss_events |>
  dplyr::filter(evid == 1, cmt %in% c("depot_sl1", "depot_q1w1", "depot_q4w1")) |>
  dplyr::select(id, time, amt, treatment, tau, ii, addl) |>
  dplyr::mutate(time = time + ii * addl) |>   # the final dose of the regimen
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "ng/mL", timeu = "hr")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "mg")

# One steady-state interval per regimen: [last dose, last dose + tau].
intervals <- ss_regimens |>
  transmute(start = tau * (ndose - 1L), end = start + tau,
            treatment = factor(label, ss_regimens$label),
            cmax = TRUE, tmax = TRUE, cav = TRUE, cmin = TRUE,
            auclast = TRUE) |>
  as.data.frame()

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)

# Per-subject metrics, with the peak-to-trough ratio derived from cmax / cmin.
gcv <- function(x) 100 * sqrt(exp(stats::var(log(x))) - 1)

nca_subj <- as.data.frame(nca_res) |>
  filter(PPTESTCD %in% c("cmax", "cav", "cmin"), !is.na(PPORRES)) |>
  select(treatment, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  mutate(ptr = cmax / cmin) |>
  tidyr::pivot_longer(c(cmax, cav, cmin, ptr),
                      names_to = "PPTESTCD", values_to = "PPORRES")
```

## Comparison against published Table 4

Table 4 reports **geometric means** (with geometric CV %), so the per-subject
PKNCA output is aggregated with the geometric mean rather than
`ncaComparisonTable()`'s default median.

```{r nca-compare}
sim_wide <- nca_subj |>
  group_by(treatment, PPTESTCD) |>
  summarise(PPORRES = gm(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  mutate(treatment = as.character(treatment))

published <- tibble::tribble(
  ~treatment,   ~cmax, ~cav,  ~cmin,  ~ptr,
  "SL 8 mg",     3.92, 1.08,  0.540,  7.26,
  "SL 16 mg",    6.09, 1.69,  0.852,  7.15,
  "SL 24 mg",    7.94, 2.21,  1.10,   7.24,
  "Q1W 8 mg",    1.69, 0.893, 0.386,  4.37,
  "Q1W 16 mg",   3.38, 1.79,  0.780,  4.34,
  "Q1W 24 mg",   4.97, 2.66,  1.18,   4.21,
  "Q1W 32 mg",   6.75, 3.57,  1.55,   4.37,
  "Q4W 64 mg",   4.11, 1.83,  0.870,  4.72,
  "Q4W 96 mg",   5.84, 2.71,  1.34,   4.36,
  "Q4W 128 mg",  8.16, 3.68,  1.74,   4.69,
  "Q4W 160 mg",  9.97, 4.52,  2.21,   4.52
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_wide,
  reference = published,
  by        = "treatment",
  units     = c(cmax = "ng/mL", cav = "ng/mL", cmin = "ng/mL",
                ptr = "Cmax,ss / Ctrough,ss"),
  tolerance_pct = 20
)

cmp |>
  mutate(treatment = factor(treatment, ss_regimens$label)) |>
  arrange(treatment) |>
  rename("Regimen" = treatment) |>
  knitr::kable(
    digits = 3,
    caption = paste("Simulated vs. published steady-state NCA (Table 4 of",
                    "Bjornsson 2023). * differs from the paper by > 20 %.")
  )
```

The paper's own geometric CV percentages are the sharpest available check on how
the Table 3 `IIV / CV (%)` column should be read (see Assumptions below), so they
are compared as well.

```{r nca-cv}
published_cv <- tibble::tribble(
  ~treatment,   ~cmax, ~cav, ~cmin, ~ptr,
  "SL 8 mg",       37,   33,    46,   46,
  "SL 16 mg",      36,   33,    45,   44,
  "SL 24 mg",      38,   35,    47,   45,
  "Q1W 8 mg",      25,   22,    45,   54,
  "Q1W 16 mg",     24,   22,    45,   53,
  "Q1W 24 mg",     25,   21,    45,   55,
  "Q1W 32 mg",     24,   21,    43,   52,
  "Q4W 64 mg",     41,   21,    40,   66,
  "Q4W 96 mg",     35,   20,    33,   52,
  "Q4W 128 mg",    40,   22,    39,   63,
  "Q4W 160 mg",    39,   22,    39,   60
) |> tidyr::pivot_longer(-treatment, names_to = "PPTESTCD", values_to = "paper_cv")

cv_cmp <- nca_subj |>
  group_by(treatment, PPTESTCD) |>
  summarise(sim_cv = gcv(PPORRES), .groups = "drop") |>
  mutate(treatment = as.character(treatment)) |>
  left_join(published_cv, by = c("treatment", "PPTESTCD")) |>
  mutate(diff_pp = sim_cv - paper_cv,
         label = dplyr::recode(PPTESTCD, cmax = "Cmax,ss", cav = "Cav,ss",
                               cmin = "Ctrough,ss", ptr = "Cmax,ss/Ctrough,ss"),
         treatment = factor(treatment, ss_regimens$label))

cv_cmp |>
  select(treatment, label, sim_cv, paper_cv, diff_pp) |>
  arrange(treatment, label) |>
  rename("Regimen" = treatment, "NCA parameter" = label,
         "Simulated geo. CV (%)" = sim_cv, "Paper geo. CV (%)" = paper_cv,
         "Difference (pp)" = diff_pp) |>
  knitr::kable(digits = 1,
               caption = paste("Geometric CV percentages, simulated vs. Table 4",
                               "of Bjornsson 2023 (pp = percentage points)."))
```

```{r nca-cv-summary}
cv_cmp |>
  group_by(`NCA parameter` = label) |>
  summarise(`Mean |difference| (pp)` = mean(abs(diff_pp)),
            `Max |difference| (pp)` = max(abs(diff_pp)), .groups = "drop") |>
  knitr::kable(digits = 2,
               caption = paste("Aggregate agreement of the simulated geometric",
                               "CV percentages with Table 4."))
```

## Which reading of the Table 3 `IIV / CV (%)` column?

Table 3 reports a single `CV (%)` column spanning both exponentially distributed
parameters and the four logit-transformed fractions. A coefficient of variation
is undefined for a logit-normal parameter, so the column can only be uniform if
it carries `100 * omega` - the standard deviation of eta on the transformed
scale. The model file adopts that reading (`variance = (CV / 100)^2`); the
alternative log-normal reading is `omega^2 = log(1 + CV^2)`.

The choice is testable, because Table 4's geometric CV percentages are a direct
readout of the omegas. The block below re-solves every regimen under the
alternative reading and scores both against Table 4. Nothing here is fitted - the
published CVs are recovered from the model's own OMEGA diagonal, so the two
matrices are two deterministic transforms of the same 11 published numbers.

```{r iiv-reading}
omega_adopted <- rxode2::rxode2(mod)$omega
cv_published  <- sqrt(diag(omega_adopted))     # recovers the Table 3 CV/100
omega_alt     <- diag(log(1 + cv_published^2))
dimnames(omega_alt) <- dimnames(omega_adopted)

# Same NCA pipeline, driven by a supplied OMEGA.
cv_under <- function(omega) {
  s <- rxode2::rxSolve(mod, events = ss_events, omega = omega,
                       keep = "treatment") |> as.data.frame()
  nc <- s |> filter(!is.na(Cc)) |> select(id, time, Cc, treatment)
  r <- PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(nc, Cc ~ time | treatment + id,
                     concu = "ng/mL", timeu = "hr"),
    dose_obj, intervals = intervals))
  as.data.frame(r) |>
    filter(PPTESTCD %in% c("cmax", "cav", "cmin"), !is.na(PPORRES)) |>
    select(treatment, id, PPTESTCD, PPORRES) |>
    tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
    mutate(ptr = cmax / cmin) |>
    tidyr::pivot_longer(c(cmax, cav, cmin, ptr),
                        names_to = "PPTESTCD", values_to = "PPORRES") |>
    group_by(treatment, PPTESTCD) |>
    summarise(sim_cv = gcv(PPORRES), .groups = "drop") |>
    mutate(treatment = as.character(treatment))
}

bind_rows(
  cv_under(omega_adopted) |> mutate(reading = "Adopted: var = (CV/100)^2"),
  cv_under(omega_alt)     |> mutate(reading = "Alternative: var = log(1 + CV^2)")
) |>
  left_join(published_cv, by = c("treatment", "PPTESTCD")) |>
  mutate(label = dplyr::recode(PPTESTCD, cmax = "Cmax,ss", cav = "Cav,ss",
                               cmin = "Ctrough,ss",
                               ptr = "Cmax,ss/Ctrough,ss")) |>
  group_by(reading, label) |>
  summarise(bias = mean(sim_cv - paper_cv),
            mae = mean(abs(sim_cv - paper_cv)), .groups = "drop") |>
  tidyr::pivot_longer(c(bias, mae), names_to = "stat", values_to = "pp") |>
  mutate(stat = dplyr::recode(stat, bias = "Mean signed diff.",
                              mae = "Mean |diff.|")) |>
  tidyr::pivot_wider(names_from = reading, values_from = pp) |>
  arrange(label, stat) |>
  rename("NCA parameter" = label, "Statistic" = stat) |>
  knitr::kable(digits = 2,
               caption = paste("Deviation (percentage points) of the simulated",
                               "geometric CV from Table 4 under the two readings",
                               "of the Table 3 IIV column, across all 11",
                               "regimens. A negative signed difference means the",
                               "simulated variability is too narrow."))
```

# Covariate effects

The final model retained age and body weight on CL, OUD status on Vc, and OUD
status and female sex on `Fq1w1`. The demographic cohort below illustrates them
for 24 mg CAM2038 Q1W at steady state. `Fq1w1` redistributes dose between the
fast and slow pathways, so it moves `Cmax` and `Ctrough` in opposite directions
while leaving overall exposure (`AUC`, hence `Cav`) unchanged - the point made in
Results 3.1.3 and Discussion 4.1.

```{r covariate-effects}
cov_arms <- tibble::tribble(
  ~label,                        ~age, ~wt, ~sexf, ~oud, ~thigh,
  "Reference (healthy M, 35 y, 72.4 kg)", 35, 72.4,  0,    0,     0,
  "Female",                               35, 72.4,  1,    0,     0,
  "OUD",                                  35, 72.4,  0,    1,     0,
  "Age 65 y",                             65, 72.4,  0,    0,     0,
  "Weight 100 kg",                        35,  100,  0,    0,     0,
  "Thigh injection",                      35, 72.4,  0,    0,     1
)

cov_events <- dplyr::bind_rows(lapply(seq_len(nrow(cov_arms)), function(i) {
  a <- cov_arms[i, ]
  make_arm("q1w", 24, 168, 12, obs_time = seq(1848, 2016, by = 2),
           label = a$label, id_offset = (i - 1L) * N_ARM,
           age = a$age, wt = a$wt, sexf = a$sexf, oud = a$oud, thigh = a$thigh)
}))
stopifnot(!anyDuplicated(unique(cov_events[, c("id", "time", "evid", "cmt")])))

sim_cov <- rxode2::rxSolve(mod, events = cov_events, keep = "treatment") |>
  as.data.frame()

sim_cov |>
  filter(!is.na(Cc)) |>
  group_by(treatment, id) |>
  summarise(cmax = max(Cc), ctrough = Cc[which.max(time)],
            cav = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2) /
              (max(time) - min(time)), .groups = "drop") |>
  group_by(treatment) |>
  summarise(across(c(cmax, cav, ctrough), gm), .groups = "drop") |>
  mutate(treatment = factor(treatment, cov_arms$label)) |>
  arrange(treatment) |>
  rename("Covariate scenario" = treatment, "Cmax,ss (ng/mL)" = cmax,
         "Cav,ss (ng/mL)" = cav, "Ctrough,ss (ng/mL)" = ctrough) |>
  knitr::kable(digits = 3,
               caption = paste("Retained covariate effects at steady state on",
                               "24 mg CAM2038 Q1W (geometric means, n = 200",
                               "per scenario)."))
```

```{r covariate-profiles}
sim_cov |>
  filter(!is.na(Cc)) |>
  group_by(treatment, time) |>
  summarise(Q50 = median(Cc), .groups = "drop") |>
  group_by(treatment) |>
  mutate(t_rel = (time - min(time)) / 24) |>
  ungroup() |>
  mutate(treatment = factor(treatment, cov_arms$label)) |>
  ggplot(aes(t_rel, Q50, colour = treatment)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time within the steady-state weekly interval (days)",
       y = "Median BPN plasma concentration (ng/mL)", colour = NULL,
       title = "Retained covariate effects, 24 mg CAM2038 Q1W at steady state") +
  theme(legend.position = "bottom")
```

# Assumptions and deviations

* **How the Table 3 `IIV / CV (%)` column is read.** Table 3 reports one
  `CV (%)` column spanning both exponentially distributed parameters (Methods
  Eq. 1) and logit-transformed parameters (`F_SL`, `F_SL1`, `Fq1w1`, `Fq4w1`;
  Methods Eqs. 2-3). A coefficient of variation is undefined for a logit-normal
  parameter, so a single column can only be uniform if it carries
  `100 * omega`, the standard deviation of eta on the transformed scale. The
  model file therefore encodes `variance = (CV / 100)^2`. The
  "Which reading of the Table 3 `IIV / CV (%)` column?" section above tests this
  against the paper's own geometric CV percentages across all 11 regimens of
  Table 4. The adopted reading wins overall, and the discriminating quantity is
  the `Cmax,ss / Ctrough,ss` ratio CV - the metric most sensitive to the large
  logit-scale etas on `Fq1w1` and `Fq4w1` - where the log-normal alternative
  `omega^2 = log(1 + CV^2)` is systematically **too narrow**. The two readings
  differ by under 1.5 % relative for the exponentially distributed parameters
  (`omega` = 0.209 versus 0.2068 for CL, for instance), so the choice materially
  affects only the four logit-transformed fractions, where the gap is larger
  (0.921 versus 0.785 for `Fq4w1`). At the 200-subjects-per-arm cohort size used
  here the margin between the two readings is modest relative to Monte Carlo
  noise on a CV; a higher-powered run of the same comparison at 1000 per arm
  (the paper's replicate count, above this vignette's cohort cap) separates them
  more cleanly, at 1.8 versus 2.7 percentage points of mean absolute error over
  the same 44 targets.
* **Where the IIV on `F_SL` enters relative to the dose effect.** Methods Eq. 2
  applies the logit transform to "the typical parameter value (P)", and Results
  3.1.3 prints the typical value as the dose-dependent
  `F_SL = 0.14 * (Dose/16)^-0.371`. The model file therefore adds eta to the
  logit of the *dose-adjusted* typical value, which keeps every individual
  `F_SL` inside (0, 1). Applying eta to the logit of the 16 mg value and then
  scaling by the dose term is numerically identical at the 16 mg reference dose
  and differs only in how variability propagates at other doses; it is not used
  here because it can push an individual `F_SL` above 1.
* **The printed `F_SL` equation is a power model, contradicting Methods Eq. 4.**
  Eq. 4 states that continuous covariates enter "as power models for
  untransformed parameters and linear models for logit-transformed parameters",
  which for the logit-transformed `F_SL` would mean an additive shift of
  `-0.371 * log(Dose/16)` on the logit scale. That form gives
  `F_SL(8 mg) = 17.4 %`, not the 18.1 % the paper itself reports. The power form
  printed in Results 3.1.3 reproduces all three published values (18.1 / 14.0 /
  12.0 % at 8 / 16 / 24 mg) exactly, so the printed equation is used and the
  Methods generalisation is treated as not applying to this term.
* **Categorical covariates on `Fq1w1` are additive on the logit scale.** Methods
  Eq. 5 describes categorical covariates as "a fractional difference to the most
  common category", which for an untransformed parameter such as `Vc` means
  `Vc * (1 + theta)` - that form reproduces the published 64.3 L (healthy) versus
  237 L (OUD) exactly and is what the model file uses. Applying the same
  multiplicative form to `logit(Fq1w1)` would make `Fq1w1` *higher* in
  participants with OUD (48.5 %) and *lower* in women (42.9 %), contradicting
  Results 3.1.3 ("Higher `Fq1w1` was shown in women than in men, and in healthy
  individuals versus those with OUD"). Adding the coefficients on the logit
  scale gives 29.9 % for OUD and 59.8 % for women, matching both stated
  directions. The additive-on-logit reading is therefore used; the paper does
  not print the `Fq1w1` covariate equation explicitly.
* **`Fq4w1` = 8.63 % (Table 3) versus 0.0900 implied in the text.** Results
  3.1.3 writes the slow-pathway Q4W fraction as "0.910 (= 1 - 0.0900)", while
  Table 3 reports `Fq4w1 = 8.63 %`. The Table 3 value is used, being the final
  model's parameter table; it reproduces the paper's "70.1 % of the CAM2038 Q4W
  dose was absorbed after 28 days" as 70.05 % (the text's 0.0900 gives 70.17 %).
* **Injection-site typo in Results 3.1.1.** The sentence "for injections in the
  thigh, the entire dose was absorbed through the slower of the two absorption
  pathways (`ka,q1w1`) with the estimated dose fraction entering compartment
  q1w1 (`Fq1w1`) close to 0" names `ka,q1w1` as the slower pathway, but
  `ka,q1w1` = 0.0401 1/h is the *faster* of the Q1W pair and `Fq1w1` = 0 routes
  the dose into `depot_q1w2` (`ka,q1w2` = 0.00565 1/h). Results 3.1.3 and
  Discussion 4.1 both confirm `ka,q1w2` is the slow pathway. The model
  implements `Fq1w1 = 0` for thigh injections, which sends the whole dose
  through `depot_q1w2`, consistent with the rest of the paper.
* **Covariates in the replication cohort are fixed, not sampled.** Methods 2.4
  states the paper's simulations were for a typical healthy 35-year-old, 72.4 kg
  male. The Table 4 / Table 5 replications hold covariates at those values so
  only inter-individual random effects vary, matching the published design. The
  covariate-effect section uses a separate scenario cohort instead of a sampled
  demographic distribution, because the paper reports covariate effects as
  point-value contrasts rather than as a distributional summary.
* **Cohort size.** 200 subjects per arm (the nlmixr2lib cap) versus the paper's
  1000 replicates. With geometric CVs of 20-66 %, the Monte Carlo standard error
  on a simulated geometric mean is roughly 1.5-4.5 % at n = 200, and on a
  simulated geometric CV roughly 1-3 percentage points. That is the floor on the
  agreement achievable in the comparison tables, and it is the reason the
  IIV-reading discriminator above is less sharp here than at the paper's n.
* **Steady state.** Dose counts (20 daily SL, 12 weekly, 10 monthly) were chosen
  from a typical-value convergence check rather than taken from the paper, which
  states only that "steady-state conditions appear to be reached within 4 doses"
  of CAM2038 and within 7 daily SL doses. At 4 weekly doses the typical-value
  `Cav,ss` is 1.6 % below its converged value, so the paper's own simulations may
  sit marginally below full steady state.
* **Tapering setup.** Table 5 describes decay "after interrupting treatment with
  CAM2038 at steady state" while Figure 7 specifies five doses. Five doses are
  used here, matching Figure 7's explicit design. Note also that Table 5's
  "Geometric CV %" column (1.33-1.82) is not a percentage - the values are
  consistent with a geometric standard deviation, not a CV, given the published
  95 % prediction intervals span 3- to 5-fold.
* **Sublingual concentrations run high; CAM2038 does not.** Simulated SL
  `Cmax,ss`, `Cav,ss`, and `Ctrough,ss` exceed Table 4 by 4-11 % at all three
  doses, all in the same direction, whereas the CAM2038 Q1W and Q4W regimens
  straddle the published values (-6.5 % to +3.6 %). The SL *shape* is right - the
  `Cmax,ss / Ctrough,ss` ratio is within 6 % and every SL geometric CV within 5
  percentage points - so the residual is a level offset on the SL bioavailable
  dose rather than a structural error in the dual SL absorption pathway. No
  reported quantity in the paper distinguishes its cause, and nothing was tuned.
  Note that the paper independently reports its own model-predicted SL
  steady-state values as running *below* the observed non-compartmental values
  (Discussion 4.1), so its published SL simulation is known to sit at the low
  end. All 44 Table 4 targets are within 12 %, well inside the 20 % flag
  threshold, so no row is starred.
* **Residual error.** Results 3.1.1 describes the RUV as additive on
  log-transformed concentrations, which is proportional on the linear scale;
  Table 3 reports it directly as `sigma_prop` = 27.7 %, encoded as
  `propSd = 0.277`.
* **No parameter uncertainty.** The simulations propagate inter-individual
  variability only. Table 3's relative standard errors are recorded in the model
  file's in-file comments but no covariance matrix is published, so parameter
  uncertainty is not propagated (the paper's own prediction intervals likewise
  reflect IIV).
* **Compartment naming.** The six absorption depots use the source paper's own
  Figure 1 labels (`depot_sl1`, `depot_sl2`, `depot_q1w1`, `depot_q1w2`,
  `depot_q4w1`, `depot_q4w2`) and are declared through the model file's
  `paper_specific_compartments` field. Four of the six name pathways of two
  proprietary CAM2038 formulations and so were not registered as new canonical
  compartments; the generic blessed `depot1` ... `depot6` numbering was rejected
  because a user dosing this four-route model must be able to tell which depot
  pair belongs to which formulation and which member is the fast pathway.
* **New canonical covariates.** `DIS_OUD` (opioid use disorder indicator) and
  `DOSE_BPN_SL_MG` (administered sublingual dose, needed because `model()` code
  cannot read the `amt` of the dose record it scales) were added to
  `inst/references/covariate-columns.md` in the same change. `INJSITE_THIGH`
  already existed and gained this model in its example list.
* **Not implemented: BLQ handling and the 192 mg Q4W dose.** Concentrations below
  the 0.0250 ng/mL LLOQ were excluded from the paper's estimation but retained
  for its visual predictive checks; the packaged model emits continuous
  concentrations and applies no BLQ rule. Dose-proportionality was reported for
  CAM2038 Q4W up to 192 mg, but Table 4 simulates only 64-160 mg, so 192 mg is
  not among the validation targets. The paper also notes a small non-linearity in
  Q4W `Cmax` at 192 mg that "was not captured by the PK model", so the packaged
  model - like the source - is linear in dose.
* **Not implemented: norbuprenorphine.** The active metabolite was measured in
  the trials but deliberately excluded from the population PK model
  (Discussion 4.1), so no metabolite compartment is present.
* **Between-occasion variability.** Not investigated by the paper ("few
  individuals received > 1 dose of CAM2038"; Discussion 4.1), so none is encoded.
* **Supplement not on disk.** Tables S1-S4 and Figures S1-S8 were not available.
  They document the trial designs, the dataset composition, the covariate
  relationships screened, and the stepwise-covariate-model results for the
  *initial* base model - none of which carry final-model parameter values. All
  values encoded here come from Table 3 and the Results narrative of the main
  paper.
