---
title: "Batiraxcept (Bonifacio 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Batiraxcept (Bonifacio 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Bonifacio_2020_batiraxcept"))$reference`
* Description: `r rxode2::rxode(readModelDb("Bonifacio_2020_batiraxcept"))$description`
* Article: <https://doi.org/10.1111/cts.12706> (open access, PMC6951457)
* Supplement: Supplementary Material `CTS-13-204-s001.docx` (model equations,
  PK and PD parameter tables, monkey VPCs)

Batiraxcept (AVB-S6-500) is a fusion of the AXL extracellular domain and a
human IgG1 Fc that binds growth arrest-specific 6 (GAS6) protein with
femtomolar affinity. Bonifacio 2020 fit a two-compartment model with parallel
linear and Michaelis-Menten (target-mediated) elimination to cynomolgus monkey
serum concentrations, scaled it allometrically to humans, and linked the
serum drug concentration to serum free GAS6 through a direct inhibitory
sigmoid Emax relationship. The typical-value human projection was used to pick
the first-in-human (FIH) doses, and the FIH study served as an external
validation.

# Population

The model was fit to data from five nonclinical cynomolgus monkey studies
covering 0.1 to 150 mg/kg: 2 animals per dose level at 0.1, 0.5 and 1 mg/kg,
4 animals at 5 mg/kg (two studies), 12 animals per dose level over
30-100 mg/kg (three studies) and 18 animals at 150 mg/kg (Methods,
"Nonclinical studies used for PK/PD model development"). The total animal
count and the animal body weights are not reported. PK parameters are
reported centred on a 70 kg subject.

The external validation cohort (NCT03401528; Table 1) was 31 healthy
volunteers receiving batiraxcept by 60-minute IV infusion: single doses of 1,
2.5, 5 or 10 mg/kg (6 per group) or 5 mg/kg once weekly for 4 weeks (7),
aged 22-54 years, median weight 75.5 kg, predominantly Black or African
American and male. The paper's pre-study simulations used a 75 kg typical
subject; this vignette does the same.

The same information is available programmatically via
`readModelDb("Bonifacio_2020_batiraxcept")()$population`.

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL at 70 kg) | 1.10 L/day | Table 2; Supplement Table 1 |
| `lvc` (VC at 70 kg) | 2.97 L | Table 2; Supplement Table 1 |
| `lvp` (VP at 70 kg) | 2.94 L | Table 2; Supplement Table 1 |
| `lq` (Q at 70 kg) | 2.04 L/day | Table 2; Supplement Table 1 |
| `lvmax` (Vmax at 70 kg) | 3.35 mg/day | Table 2; Supplement Table 1 |
| `lkm` (KM) | 102 ng/mL = 0.102 ug/mL | Table 2; Supplement Table 1 |
| `e_wt_cl` | 0.75 (not estimated) | Methods "Model predictions of clinical PK/PD"; Supplement "Scaling to Human Pharmacokinetics" |
| `e_wt_vc` | 1.0 (not estimated) | Same as above |
| `lrbase` (E0, baseline GAS6) | 23.8 ng/mL | Supplement Table 2 |
| `lec50` (EC50) | 21.4 ng/mL = 0.0214 ug/mL | Supplement Table 2 |
| `lhill` (H) | 0.796 | Supplement Table 2 |
| `etalcl`, `etalvc`, `etalvp`, `etalq`, `etalvmax`, `etalkm` | 29.7, 14.0, 14.3, 50.0, 98.6, 103 CV% -> omega^2 = log(CV^2 + 1) | Table 2; Supplement Table 1 |
| `propSd` | proportional; magnitude not reported, encoded `fixed(0)` | Supplement "Non-Human Primate Pharmacokinetic Analysis" |
| `d/dt(central)`, `d/dt(peripheral1)` | two-compartment with parallel linear + Michaelis-Menten loss of `Ap` | Supplement Methods (dAp/dt, dAt/dt); Figure 1 |
| `GAS6 <- E0 * (1 - C^H / (EC50^H + C^H))` | direct effect | Supplement "Non-Human Primate Pharmacodynamic Analysis" |
| Allometric scaling `(WT/70)^0.75` on CL, Q, Vmax; `(WT/70)^1` on VC, VP | | Supplement "Scaling to Human Pharmacokinetics" |

# Typical-subject simulation of the FIH regimens

The paper's human projections (Figures 2 and 3, Table 3) are typical-value
predictions for a 75 kg subject; between-animal variability was deliberately
not propagated (Supplement Results; Discussion). The random effects are
therefore zeroed here.

```{r events}
wt_sim <- 75
inf_dur <- 1 / 24 # 60-minute infusion, in days

regimens <- tibble::tribble(
  ~treatment,           ~dose_mgkg, ~n_doses,
  "1 mg/kg",            1,          1L,
  "2.5 mg/kg",          2.5,        1L,
  "5 mg/kg",            5,          1L,
  "10 mg/kg",           10,         1L,
  "5 mg/kg q.w. x 4",   5,          4L
)

obs_times <- sort(unique(c(
  seq(0, 63, by = 1 / 24),
  inf_dur + 7 * (0:3)
)))

make_regimen <- function(i) {
  r <- regimens[i, ]
  dose_times <- 7 * (seq_len(r$n_doses) - 1)
  doses <- tibble(
    id = i, time = dose_times, evid = 1L, amt = r$dose_mgkg * wt_sim,
    rate = r$dose_mgkg * wt_sim / inf_dur, cmt = "central"
  )
  obs <- tibble(
    id = i, time = obs_times, evid = 0L, amt = 0, rate = 0, cmt = "central"
  )
  bind_rows(doses, obs) |>
    mutate(treatment = r$treatment, WT = wt_sim) |>
    arrange(time, desc(evid))
}

events_typ <- bind_rows(lapply(seq_len(nrow(regimens)), make_regimen))
stopifnot(!anyDuplicated(events_typ[, c("id", "time", "evid")]))
```

```{r simulate-typical}
mod <- readModelDb("Bonifacio_2020_batiraxcept")
mod_typ <- rxode2::zeroRe(mod)
sim_typ <- rxode2::rxSolve(
  mod_typ,
  events = events_typ, keep = c("treatment"),
  rtol = 1e-10, atol = 1e-12
) |>
  as.data.frame() |>
  mutate(treatment = factor(treatment, levels = regimens$treatment))
```

## Figure 2: serum batiraxcept

```{r figure-2}
# Replicates the model lines of Figure 2 of Bonifacio 2020 (typical 75 kg
# subject). LLOQ 10 ng/mL = 0.01 ug/mL; values below LLOQ plotted at LLOQ/2
# as in the paper.
lloq_pk <- 0.01
sim_typ |>
  filter(time > 0) |>
  mutate(Cc_plot = ifelse(Cc < lloq_pk, lloq_pk / 2, Cc)) |>
  ggplot(aes(time, Cc_plot)) +
  geom_line(colour = "steelblue") +
  geom_hline(yintercept = lloq_pk, linetype = "dashed", colour = "red") +
  facet_wrap(~treatment) +
  scale_y_log10() +
  labs(
    x = "Time (day)", y = "Batiraxcept (ug/mL)",
    caption = "Replicates the typical-subject predictions of Figure 2 of Bonifacio 2020."
  )
```

## Figure 3: serum GAS6

```{r figure-3}
# Replicates the model lines of Figure 3 of Bonifacio 2020. GAS6 LLOQ in the
# FIH study was 2 ng/mL; values below LLOQ plotted at LLOQ/2.
lloq_gas6 <- 2
sim_typ |>
  mutate(GAS6_plot = ifelse(GAS6 < lloq_gas6, lloq_gas6 / 2, GAS6)) |>
  ggplot(aes(time, GAS6_plot)) +
  geom_line(colour = "darkgreen") +
  geom_hline(yintercept = lloq_gas6, linetype = "dashed", colour = "red") +
  facet_wrap(~treatment) +
  scale_y_log10() +
  labs(
    x = "Time (day)", y = "Serum GAS6 (ng/mL)",
    caption = "Replicates the typical-subject predictions of Figure 3 of Bonifacio 2020."
  )
```

# PKNCA validation against Table 3 (model-predicted values)

Table 3 of Bonifacio 2020 lists the model-predicted Cmax and AUC0-tau for the
typical 75 kg subject next to the observed FIH values. The predicted column is
the direct reproduction target for this model; the observed column is an
external comparison that the paper itself reports as underpredicted for AUC
(44.6-72.4% predicted/observed). AUC0-tau is computed over one week
(0-7 days for the single doses and the first weekly dose, 21-28 days for the
fourth weekly dose), which is the interval that reproduces the paper's
predicted values. Model time is in days, so the published ug*h/mL values are
divided by 24.

```{r pknca}
conc_df <- sim_typ |>
  filter(!is.na(Cc)) |>
  mutate(Cc = pmax(Cc, 0)) |>
  select(id, time, Cc, treatment) |>
  mutate(treatment = as.character(treatment))

dose_df <- events_typ |>
  filter(evid == 1) |>
  select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- bind_rows(
  data.frame(
    treatment = regimens$treatment, start = 0, end = 7,
    cmax = TRUE, auclast = TRUE
  ),
  data.frame(
    treatment = "5 mg/kg q.w. x 4", start = 21, end = 28,
    cmax = TRUE, auclast = TRUE
  )
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)

sim_nca <- as.data.frame(nca_res$result) |>
  mutate(
    group = case_when(
      treatment == "5 mg/kg q.w. x 4" & start == 0 ~ "5 mg/kg q.w. x 4, dose 1",
      treatment == "5 mg/kg q.w. x 4" & start == 21 ~ "5 mg/kg q.w. x 4, dose 4",
      TRUE ~ treatment
    )
  ) |>
  select(group, PPTESTCD, PPORRES)

# Table 3, predicted values (Cmax ug/mL; AUC0-tau ug*h/mL converted to ug*day/mL)
published <- tibble::tribble(
  ~group,                      ~cmax, ~auclast,
  "1 mg/kg",                   23.2,  869 / 24,
  "2.5 mg/kg",                 58.2,  2520 / 24,
  "5 mg/kg",                   116,   5290 / 24,
  "10 mg/kg",                  233,   10800 / 24,
  "5 mg/kg q.w. x 4, dose 1",  116,   5290 / 24,
  "5 mg/kg q.w. x 4, dose 4",  134,   7060 / 24
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_nca,
  reference = published,
  by = "group",
  units = c(cmax = "ug/mL", auclast = "ug*day/mL"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  digits = 2,
  caption = "Simulated typical-subject NCA vs. Bonifacio 2020 Table 3 model-predicted values. * differs by >20%."
)
```

```{r pknca-gate}
chk <- sim_nca |>
  inner_join(
    published |> pivot_longer(-group, names_to = "PPTESTCD", values_to = "ref"),
    by = c("group", "PPTESTCD")
  ) |>
  mutate(pct_diff = 100 * (PPORRES / ref - 1))
knitr::kable(chk, digits = 2)
# Both sides are typical-value predictions of the same model, so the only
# differences are the paper's rounding and its (unreported) output grid and
# integration. Measured: Cmax -0.6% to -1.0%, AUC +1.0% to +1.9%. A 5% bound
# still fails on any mis-transcribed volume, clearance, Vmax or unit, each of
# which moves these values by tens of percent.
stopifnot(nrow(chk) == 12L, all(abs(chk$pct_diff) < 5))
```

# GAS6 suppression versus the paper's narrative

The Results describe the typical-subject GAS6 prediction qualitatively at
each follow-up visit (GAS6 LLOQ 2 ng/mL, baseline 23.8 ng/mL). The table below
shows the simulated typical GAS6 at those visits.

```{r gas6-visits}
gas6_visits <- sim_typ |>
  filter(treatment != "5 mg/kg q.w. x 4", abs(time - round(time)) < 1e-6, round(time) %in% c(7, 14, 21, 28)) |>
  mutate(time = round(time)) |>
  select(treatment, time, Cc, GAS6) |>
  mutate(pct_of_baseline = 100 * GAS6 / 23.8)
knitr::kable(gas6_visits, digits = 3)

g <- function(trt, day) gas6_visits$GAS6[gas6_visits$treatment == trt & gas6_visits$time == day]
stopifnot(
  # "undetectable" at day 7 for every single dose, and at day 14 for 5 and
  # 10 mg/kg (paper Results).
  g("1 mg/kg", 7) < lloq_gas6,
  g("2.5 mg/kg", 7) < lloq_gas6,
  g("5 mg/kg", 14) < lloq_gas6,
  g("10 mg/kg", 14) < lloq_gas6,
  # "nearly baseline" at day 14 for 1 mg/kg and day 21 for 5 mg/kg, and at
  # day 28 for 10 mg/kg (above half of baseline).
  g("1 mg/kg", 14) > 0.5 * 23.8,
  g("5 mg/kg", 21) > 0.5 * 23.8,
  g("10 mg/kg", 28) > 0.5 * 23.8,
  # "just-detectable" at day 14 for 2.5 mg/kg and day 21 for 10 mg/kg: above
  # the LLOQ but well below baseline.
  g("2.5 mg/kg", 14) > lloq_gas6, g("2.5 mg/kg", 14) < 0.5 * 23.8,
  g("10 mg/kg", 21) > lloq_gas6, g("10 mg/kg", 21) < 0.5 * 23.8
)
```

The Supplement also states that batiraxcept concentrations above
~2000 ng/mL suppress GAS6 below the monkey assay LLOQ of 0.78 ng/mL, and that a
concentration at the 10 ng/mL PK LLOQ corresponds to at least 50% suppression.
Checking both against the PD equation:

```{r pd-statements}
pd <- function(C) 23.8 * (1 - C^0.796 / (0.0214^0.796 + C^0.796))
c(GAS6_at_2_ug_per_mL = pd(2), GAS6_at_0.01_ug_per_mL = pd(0.01))
stopifnot(pd(2) < 0.78)
```

With the printed estimates, GAS6 at 2 ug/mL is below the 0.78 ng/mL monkey
LLOQ, as stated. At the 10 ng/mL PK LLOQ, however, GAS6 is only
`r round(100 * (1 - pd(0.01) / 23.8))`% suppressed, not "at least 50%":
EC50 (21.4 ng/mL) is above the PK LLOQ, so 50% suppression is reached only at
21.4 ng/mL. This is an approximation in the paper's wording and was not tuned.

# Between-subject variability (illustration)

The paper states that, at low doses, the TMDD variability (Vmax and KM
CV ~100%) produces a wide prediction interval while at 5-10 mg/kg the pathway
is saturated and variability matters less. The between-animal variances are
packaged for completeness, but the authors note they have no known
relationship to human variability and did not use them for human projections.
The simulation below (200 virtual 75 kg subjects per dose, single dose, no
residual error because its magnitude was not reported) illustrates the
statement.

```{r bsv}
rxode2::rxSetSeed(20200101)
bsv_regs <- regimens |> filter(n_doses == 1L)
events_bsv <- bind_rows(lapply(seq_len(nrow(bsv_regs)), function(i) {
  r <- bsv_regs[i, ]
  ids <- (i - 1L) * 200L + seq_len(200L)
  d <- tibble(
    id = ids, time = 0, evid = 1L, amt = r$dose_mgkg * wt_sim,
    rate = r$dose_mgkg * wt_sim / inf_dur, cmt = "central"
  )
  o <- tidyr::expand_grid(id = ids, time = seq(0, 28, by = 0.25)) |>
    mutate(evid = 0L, amt = 0, rate = 0, cmt = "central")
  bind_rows(d, o) |>
    mutate(treatment = r$treatment, WT = wt_sim) |>
    arrange(id, time, desc(evid))
}))
stopifnot(!anyDuplicated(events_bsv[, c("id", "time", "evid")]))

sim_bsv <- rxode2::rxSolve(mod, events = events_bsv, keep = "treatment") |>
  as.data.frame() |>
  mutate(treatment = factor(treatment, levels = bsv_regs$treatment))

sim_bsv |>
  filter(time > 0) |>
  group_by(treatment, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = pmax(Q05, 1e-3), ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue") +
  geom_hline(yintercept = lloq_pk, linetype = "dashed", colour = "red") +
  facet_wrap(~treatment) +
  scale_y_log10() +
  labs(
    x = "Time (day)", y = "Batiraxcept (ug/mL)",
    caption = "Median and 90% interval, between-animal variability applied to 75 kg subjects (illustrative)."
  )
```

```{r bsv-spread}
spread <- sim_bsv |>
  filter(abs(time - 14) < 1e-6) |>
  group_by(treatment) |>
  summarise(
    frac_below_lloq = mean(Cc < lloq_pk),
    iqr_log10 = diff(quantile(log10(pmax(Cc, 1e-6)), c(0.25, 0.75))),
    .groups = "drop"
  )
knitr::kable(spread, digits = 2)
```

# Assumptions and deviations

- **Drug name.** The paper refers only to the development code AVB-S6-500;
  the file uses the INN batiraxcept.
- **Residual error.** The Supplement states that residual variability was
  proportional but does not report its magnitude (in either the main text or
  the Supplement), so `propSd` is encoded as `fixed(0)`. Simulations with
  residual error need a user-supplied value.
- **Omega scale and covariances.** Between-animal variability is reported as
  CV%; it was converted with omega^2 = log(CV^2 + 1) (the paper does not state
  which CV convention it used). The Supplement says "variance terms were
  modeled with variance-covariance interactions" but the covariances are not
  reported, so a diagonal omega is used. These variances describe monkeys and
  were not used by the authors for human projections.
- **PD model.** The GAS6 Emax model was fit separately (Supplement Table 2
  reports estimate / SE / statistic / p-value, i.e. a regression fit, not a
  joint NONMEM fit), with no residual-error or between-subject variability
  reported. It is packaged as the algebraic output `GAS6` (ng/mL) driven by
  `Cc` (ug/mL; EC50 converted from 21.4 ng/mL) and carries no error model.
  No interspecies scaling was applied to the PD (paper Methods), so E0 is the
  monkey baseline (23.8 ng/mL).
- **LLOQ statements.** The Supplement's "PK LLOQ implies at least 50%
  suppression" does not hold exactly with the printed EC50 (see above); this
  is recorded, not tuned.
- **External validation.** Observed FIH Cmax and AUC (Table 3) are not
  gated: the paper reports AUC predicted/observed of 44.6-72.4%, i.e. the
  model (as published) underpredicts human exposure at later times.
- **Monkey population.** Animal weights and the total animal count are not
  reported, so the monkey fit cannot be re-simulated at animal-level
  covariates; the monkey VPCs (Supplement Figures 2 and 3) are not
  reproduced.
