---
title: "Vinyl chloride PBPK (Boone 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vinyl chloride PBPK (Boone 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(nlmixr2lib::readModelDb("Boone_2025_vinylchloride_pbpk"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Article: <https://doi.org/10.3390/jox15020042>\n")
cat("* Supplement (Berkeley Madonna listing): ",
    "<https://www.mdpi.com/2039-4713/15/2/42/s1>\n", sep = "")
```

* Description: `r ui$description`

Boone et al. applied the Agency for Toxic Substances and Disease Registry
(ATSDR) human PBPK toolkit to reconstruct community exposure after the 2012
Paulsboro, New Jersey train derailment, in which a punctured tanker car
released roughly 24,000 gallons of vinyl chloride (VC). The toolkit model is a
generic volatile-organic-compound (VOC) whole-body PBPK structure, recoded by
ATSDR from the Clewell et al. VC model onto Berkeley Madonna 10.4.2 and
parameterised with VC-specific partition coefficients and metabolic constants.
The complete source listing is reproduced verbatim in the paper's
Supplementary Materials under the heading *"Berkeley Madonna code for PBPK
model of Vinyl Chloride"*, and every value in the packaged model file is taken
from that listing.

## Structure

Seven flow-limited, well-mixed tissue compartments carry the chemical
(`a_venous`, `a_rapidly_perfused`, `a_slowly_perfused`, `a_fat`, `a_liver`,
`a_kidney`, `a_skin`). Arterial blood is **not** a state: the lung is assumed
to equilibrate instantaneously, so the arterial concentration is the algebraic
steady-state mixing of returning venous blood with inhaled air across the
blood:air partition coefficient,

$$C_A = \frac{Q_C\,C_V + Q_P\,C_I}{Q_C + Q_P/P_B}.$$

This is why inhalation exposure enters the model as an **air concentration**
rather than as a dose record: there are no dosing events anywhere in this
vignette. Hepatic metabolism is saturable (Michaelis-Menten on the liver
capillary concentration), and six further states integrate cumulative
metabolised, inhaled, exhaled, ingested and dermally transferred amounts so
that the listing's mass balance can be checked.

Three exposure routes are implemented simultaneously, exactly as in the
listing: inhalation of a square-wave air concentration in ppm, repeated oral
ingestion of contaminated drinking water, and dermal transfer from water
across the skin. Every published simulation in the paper uses inhalation only
(`dose_oral = 0`, `conc_water = 0`, `skin_time = 0`), because "the majority of
exposure occurred via inhalation" (Methods 2.2).

# Population

```{r population, echo = FALSE}
pop <- ui$population
tibble::tibble(Field = names(pop),
               Value = vapply(pop, function(x) paste(format(x), collapse = "; "),
                              character(1))) |>
  knitr::kable(caption = "Population metadata carried by the model file.")
```

This is a deterministic exposure-reconstruction model, not a fit: no
individual-level kinetic data were collected in Paulsboro, so the model carries
**no between-subject variability and no residual error**. Every parameter is a
recoded literature constant and is therefore wrapped in `fixed()` in `ini()`.
The Berkeley Madonna listing hard-codes a single 70 kg reference adult
(`BW = 70.0`, cited to Clewell et al. 2000); body weight is carried here as the
covariate `WT` so the model can be re-scaled, since all seven tissue volumes
are linear in body weight and cardiac output, alveolar ventilation and the
metabolic `vmax` are allometric at exponent 0.75.

The paper handles interindividual toxicokinetic variability **outside** the
model, by applying an assessment factor of 3.16 (the square root of 10) to the
predicted peak blood levels.

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Boone_2025_vinylchloride_pbpk.R` carries an in-file
comment naming its Berkeley Madonna symbol and the literature attribution
printed alongside it in the listing. The table below collects them.

| Model parameter | Berkeley Madonna symbol | Value | Attribution in the listing |
|---|---|---|---|
| `qp_per_kg` | `QPC` | 24 | Covington et al., 2007 |
| `qc_per_kg` | `QCC` | 16.5 | Clewell et al., 2000 |
| `e_wt_flow` | exponent in `QC=QCC*BW**.75` | 0.75 | listing |
| `fv_fat` | `VFC` | 0.214 | Clewell et al., 2000 |
| `fv_liver` | `VLC` | 0.026 | Clewell et al., 2000 |
| `fv_blood` | `VBloodC` | 0.079 | listing |
| `fv_rapidly_perfused` | `VRC` | 0.09 | listing |
| `fv_slowly_perfused` | `VSC` | 0.82 | listing |
| `fv_skin` | `VSkC` | 0.051 | listing |
| `fv_kidney` | `VKC` | 0.044 | listing |
| `fq_fat` | `QFC` | 0.052 | Clewell et al., 2000 |
| `fq_liver` | `QLC` | 0.24 | Fisher, Mahle and Abbas, 1998 |
| `fq_rapidly_perfused` | `QRC` | 0.7 | Clewell, 2005 |
| `fq_slowly_perfused` | `QSC` | 0.30 | Clewell, 2005 |
| `fq_kidney` | `QKC` | 0.197 | listing |
| `fq_skin` | `QSkC` | 0.05 | listing |
| `sa_skin` | `SA` | 19975 cm^2 | listing |
| `pc_blood_air` | `PB` | 1.16 | Clewell, 2001 |
| `pc_liver` | `PL` | 1.45 | Clewell, 2001 |
| `pc_fat` | `PF` | 20.7 | Clewell, 2001 |
| `pc_rapidly_perfused` | `PR` | 1.45 | Clewell, 2001 |
| `pc_slowly_perfused` | `PS` | 0.83 | Clewell, 2001 |
| `pc_kidney` | `PK` | 1.45 | listing, "(see liver)" |
| `pc_skin` | `PSk` | 1.45 | Poet et al., 2000 |
| `mw` | `MW` | 62.5 g/mol | EPA 2000 |
| `vmax_per_kg` | `Vmaxc` | 3.97 | Reitz, 1996 |
| `km` | `Km` | 0.04 mg/L | Reitz, 1996 |
| `molar_vol` | 24450 in `CIX=CONC*MW/24450` | 24450 mL/mol | listing |
| `f_alveolar` | 0.7 in `CXppm=(.7*CX+.3*CI)*24450/MW` | 0.7 | listing |
| `perm_skin` | `Kp` | 0.015 cm/h | Poet et al., 2000 |
| `pc_skin_water` | `PSkliq` | 53 | Poet et al., 2000 |
| `conc_air_ppm` | `CONC` | 0 (scenario input) | listing default |
| `inhale_time` | `inhale_time` | 5 h (scenario input) | listing default |
| `inhale_interval` | `inhale_interval` | 100000 h | listing default |
| `dose_oral` | `pdose` | 0 | listing default |
| `drink_time` | `drink_time` | 0.25 h | listing |
| `drink_interval` | `drink_interval` | 6 h | listing |
| `conc_water` | `Cliq` | 0 | listing default, "Dr. Fisher" |
| `skin_time` | `skin_time` | 0 | listing default |
| `skin_interval` | `skin_interval` | 100000 h | listing default |

Structural equations are transcribed one-for-one from the listing's
`{Model Equations}` block; each `model()` line carries the originating
Berkeley Madonna statement as a trailing comment. Two encodings differ in form
but not in value:

* Berkeley Madonna's `MOD(TIME, P)` is written as
  `time - P * floor(time / P)`, and its `IF ... THEN 0 ELSE 1` square wave as
  the complement of the comparison, which rxode2 evaluates to 1 or 0.
* The listing's separate net-absorption state `InhDOSE' = QP*(CI-CX)` is
  derived algebraically as `Ainhaled_net <- a_inhaled - a_exhaled` rather than
  carried as a redundant fourteenth state; the two are identical by
  construction.

# Exposure scenarios

Because exposure is parameter-driven rather than dose-driven, a scenario is
defined by an air concentration (`conc_air_ppm`) and an exposure-window length
(`inhale_time`), and the event table contains **observation rows only**. The
helper below runs one scenario. Observations are placed on the `a_venous` ODE
state; rxode2 returns every algebraic observable (`Cc`, `Cvenous`, `Cexhaled`,
...) as a column at those rows.

```{r sim-helper}
mod <- nlmixr2lib::readModelDb("Boone_2025_vinylchloride_pbpk")

#' Simulate one inhalation-exposure scenario.
#'
#' @param air_ppm air concentration during the exposure window (ppm)
#' @param window_h length of the exposure window (h)
#' @param tmax end of the simulation (h); defaults to the window length
#' @param dt output grid (h); the Berkeley Madonna listing used dtout = 0.01
simulate_exposure <- function(air_ppm, window_h, tmax = window_h, dt = 0.01) {
  times <- sort(unique(c(seq(0, tmax, by = dt), window_h)))
  events <- data.frame(
    id = 1L, time = times, evid = 0L,
    amt = NA_real_, cmt = "a_venous"
  )
  out <- rxode2::rxSolve(
    mod, events,
    params = c(
      WT = 70, conc_air_ppm = air_ppm,
      inhale_time = window_h, inhale_interval = 1e5
    ),
    returnType = "data.frame"
  )
  # rxSolve omits `id` for a single-subject solve; restore it.
  if (is.null(out$id)) out$id <- 1L
  out
}
```

Two readouts are used below and the distinction matters, because it is what
explains the one internally inconsistent row of the paper's Table 4:

* **peak during exposure** -- the maximum over output times strictly inside the
  exposure window, i.e. while the air switch `air_on` is still 1;
* **immediately after the window closes** -- the first output time at or after
  `inhale_time`, where `air_on` has dropped to 0 and the inhaled-air term
  vanishes from the arterial equation.

```{r readouts}
peak_in_window <- function(sim, window_h, col) max(sim[[col]][sim$time < window_h])
first_after_window <- function(sim, window_h, col) sim[[col]][sim$time >= window_h][1]
```

# Replicating Figure 2

Figure 2 of Boone et al. shows predicted arterial blood concentrations over
four hours of continuous exposure at the air levels measured shortly after the
derailment. Supplementary Table S2 reports the peak VC values converted from
the VOC monitoring data: 1444 ppm at North Commerce and 444.6 ppm at North
Delaware (outside the evacuation area). The AEGL-2 60-minute level of 1200 ppm
that the Results narrative singles out is shown alongside them.

```{r figure-2}
fig2_levels <- tibble::tribble(
  ~label,                              ~air_ppm,
  "N Commerce peak (Table S2)",          1444.0,
  "AEGL-2, 60 min (Table S5)",           1200.0,
  "N Delaware peak (Table S2)",           444.6
)

fig2 <- fig2_levels |>
  rowwise() |>
  reframe(
    label = label, air_ppm = air_ppm,
    simulate_exposure(air_ppm, window_h = 4, tmax = 4) |>
      select(time, Cc)
  )

ggplot(fig2, aes(time, Cc, colour = factor(air_ppm, levels = fig2_levels$air_ppm,
                                           labels = fig2_levels$label))) +
  geom_line(linewidth = 0.9) +
  labs(
    x = "Time (h)", y = "Arterial blood VC (mg/L)", colour = NULL,
    title = "Figure 2 - predicted arterial blood over 4 h of exposure",
    caption = "Replicates Figure 2 of Boone et al. 2025."
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The Results narrative states that at the AEGL-2 concentration of approximately
1200 ppm "the model estimated the arterial blood concentration to be around
3.17 mg/L after an hour of exposure", with "slight increases in concentration
over time as exposure continued". Both features reproduce:

```{r figure-2-anchor}
anchor <- simulate_exposure(1200, window_h = 4, tmax = 4)
tibble::tibble(
  `Exposure duration (h)` = c(1, 2, 3, 4),
  `Simulated CA (mg/L)` = vapply(
    c(1, 2, 3, 4),
    function(h) anchor$Cc[which.min(abs(anchor$time - h))],
    numeric(1)
  )
) |>
  knitr::kable(digits = 3,
               caption = "Arterial blood at 1200 ppm; the paper reports ~3.17 mg/L at 1 h.")
```

# Replicating Table 4 (AEGL blood and breath levels)

Table 4 tabulates predicted maximum arterial blood (CA), venous blood (CV) and
exhaled breath (CX) concentrations at each of the three Acute Exposure
Guideline Levels for VC over five exposure durations. The air levels are
Table S5.

```{r table-4-run}
table4_published <- tibble::tribble(
  ~aegl,     ~window_h, ~air_ppm, ~CA_pub, ~CV_pub, ~CX_pub,
  "AEGL-1",      1 / 6,      450,    0.99,    0.57,    0.86,
  "AEGL-1",      0.5,        310,    0.72,    0.47,    0.62,
  "AEGL-1",      1,          250,    0.60,    0.42,    0.52,
  "AEGL-1",      4,          140,    0.11,    0.25,    0.09,
  "AEGL-1",      8,           70,    0.17,    0.13,    0.15,
  "AEGL-2",      1 / 6,     2800,    6.81,    4.93,    5.87,
  "AEGL-2",      0.5,       1600,    4.11,    3.32,    3.55,
  "AEGL-2",      1,         1200,    3.18,    2.70,    2.74,
  "AEGL-2",      4,          820,    0.82,    1.85,    0.71,
  "AEGL-2",      8,          820,    2.19,    1.89,    1.89,
  "AEGL-3",      1 / 6,    12000,   29.68,   22.28,   25.59,
  "AEGL-3",      0.5,       6800,   18.04,   15.37,   15.55,
  "AEGL-3",      1,         4800,   13.23,   11.98,   11.41,
  "AEGL-3",      4,         3400,    3.97,    8.95,    3.42,
  "AEGL-3",      8,         3400,    9.65,    9.11,    8.32
)

table4 <- table4_published |>
  rowwise() |>
  mutate(
    .sim = list(simulate_exposure(air_ppm, window_h, tmax = window_h + 0.05)),
    CA_peak = peak_in_window(.sim, window_h, "Cc"),
    CV_peak = peak_in_window(.sim, window_h, "Cvenous"),
    CX_peak = peak_in_window(.sim, window_h, "Cexhaled"),
    CA_post = first_after_window(.sim, window_h, "Cc"),
    CX_post = first_after_window(.sim, window_h, "Cexhaled")
  ) |>
  ungroup() |>
  select(-.sim) |>
  mutate(
    duration = c("10 min", "30 min", "60 min", "4 h", "8 h")[
      match(window_h, c(1 / 6, 0.5, 1, 4, 8))
    ],
    dCA = 100 * (CA_peak - CA_pub) / CA_pub,
    dCV = 100 * (CV_peak - CV_pub) / CV_pub,
    dCX = 100 * (CX_peak - CX_pub) / CX_pub
  )

table4 |>
  select(aegl, duration, air_ppm,
         CA_pub, CA_peak, dCA, CV_pub, CV_peak, dCV, CX_pub, CX_peak, dCX) |>
  dplyr::rename(
    "AEGL" = aegl, "Duration" = duration, "Air (ppm)" = air_ppm,
    "CA pub" = CA_pub, "CA sim" = CA_peak, "CA %diff" = dCA,
    "CV pub" = CV_pub, "CV sim" = CV_peak, "CV %diff" = dCV,
    "CX pub" = CX_pub, "CX sim" = CX_peak, "CX %diff" = dCX
  ) |>
  knitr::kable(
    digits = 2,
    caption = paste(
      "Table 4 of Boone et al. reproduced with the peak-during-exposure readout.",
      "The three 4-hour rows are addressed below."
    )
  )
```

## The 4-hour rows are a readout artifact, not a model discrepancy

In the published Table 4 the 4-hour rows report an arterial concentration
**below** the venous concentration (AEGL-1: CA 0.11 vs CV 0.25; AEGL-2: 0.82 vs
1.85; AEGL-3: 3.97 vs 8.95). That ordering is impossible while a subject is
still inhaling: during uptake the arterial blood is the mixture of returning
venous blood with inhaled air, so `CA > CV` necessarily. It becomes possible
only *after* exposure stops, when the inhaled-air term disappears and arterial
blood becomes venous blood diluted by clean air across the lung.

Reading CA and CX from the first output step **after** the window closes -
while still reading CV at the end of the window - recovers the published
numbers to within a fraction of a percent:

```{r table-4-artifact}
table4 |>
  filter(window_h == 4) |>
  transmute(
    AEGL = aegl,
    `Air (ppm)` = air_ppm,
    `CA pub` = CA_pub,
    `CA sim (post-window)` = CA_post,
    `CA %diff` = 100 * (CA_post - CA_pub) / CA_pub,
    `CX pub` = CX_pub,
    `CX sim (post-window)` = CX_post,
    `CX %diff` = 100 * (CX_post - CX_pub) / CX_pub,
    `CV pub` = CV_pub,
    `CV sim (window end)` = CV_peak,
    `CV %diff` = dCV
  ) |>
  knitr::kable(
    digits = 2,
    caption = paste(
      "The 4-hour rows of Table 4 read CA and CX one output step after the",
      "exposure window closed, while CV was read at the end of the window."
    )
  )
```

```{r table-4-artifact-check, echo = FALSE}
art <- table4 |> filter(window_h == 4)
stopifnot(nrow(art) == 3L)
# Post-window readout reproduces the published 4-hour CA to better than 1.2%.
stopifnot(all(abs(100 * (art$CA_post - art$CA_pub) / art$CA_pub) < 1.2))
# CX likewise, except for the AEGL-1 row, whose published 0.09 mg/L carries
# ~1% per unit in its last printed decimal; that row alone sits at 6.6%.
stopifnot(all(abs(100 * (art$CX_post - art$CX_pub) / art$CX_pub) < 7))
stopifnot(sum(abs(100 * (art$CX_post - art$CX_pub) / art$CX_pub) < 0.2) == 2L)
# ... whereas the peak-during-exposure readout is >100% high, confirming the
# published rows cannot have come from the in-window readout used elsewhere.
stopifnot(all(100 * (art$CA_peak - art$CA_pub) / art$CA_pub > 100))
```

## Agreement on the remaining twelve rows

```{r table-4-summary}
table4 |>
  filter(window_h != 4) |>
  summarise(
    `Rows` = n(),
    `max |CA %diff|` = max(abs(dCA)),
    `max |CV %diff|` = max(abs(dCV)),
    `max |CX %diff|` = max(abs(dCX))
  ) |>
  knitr::kable(digits = 2,
               caption = "Worst-case agreement across the twelve non-4-hour rows.")
```

```{r table-4-assert, echo = FALSE}
ok <- table4 |> filter(window_h != 4)
stopifnot(nrow(ok) == 12L)
stopifnot(all(abs(ok$dCA) < 2.5), all(abs(ok$dCX) < 2.5), all(abs(ok$dCV) < 7))

# Tier A -- the 30- and 60-minute rows, where every published value has enough
# printed precision to be a sharp test: better than 0.7% on all three outputs.
tierA <- ok |> filter(window_h %in% c(0.5, 1))
stopifnot(nrow(tierA) == 6L)
stopifnot(all(abs(tierA$dCA) < 0.7), all(abs(tierA$dCV) < 0.7),
          all(abs(tierA$dCX) < 0.7))

# Tier B -- the 8-hour rows. Bounded at 2.2% because the AEGL-1 row publishes
# CA/CV/CX as 0.17/0.13/0.15, where one unit in the last printed decimal is
# already 0.6-0.8%.
tierB <- ok |> filter(window_h == 8)
stopifnot(nrow(tierB) == 3L)
stopifnot(all(abs(tierB$dCA) < 2.2), all(abs(tierB$dCV) < 2.2),
          all(abs(tierB$dCX) < 2.2))

# Tier C -- the 10-minute rows: CA and CX within 2.3%, but CV is consistently
# 6-7% low across all three AEGL tiers (see Errata). The sign is asserted too,
# because a consistent one-sided offset is the signature of a readout
# difference rather than of numerical noise.
tierC <- ok |> filter(window_h < 0.5)
stopifnot(nrow(tierC) == 3L)
stopifnot(all(abs(tierC$dCA) < 2.3), all(abs(tierC$dCX) < 2.3))
stopifnot(all(tierC$dCV < -6), all(tierC$dCV > -7))
```

The six 30-minute and 60-minute rows -- those whose published values carry
enough printed precision to be a sharp test -- agree with the publication to
better than **0.7%** on all three outputs. The three 8-hour rows agree to
within 2.2%, the residual being dominated by the AEGL-1 row, which publishes
CA, CV and CX as 0.17, 0.13 and 0.15 mg/L, where a single unit in the last
printed decimal is already most of that difference. The three 10-minute rows
agree to within 2.3% on CA and CX but sit 6.7-6.9% below the published CV; see
Errata.

# Replicating Table 3 (Paulsboro reconstruction)

Table 3 predicts arterial blood concentrations for the measured air levels on
each day of the incident, inside and outside the evacuation area, "after 24 h
of exposure". The air concentrations are Supplementary Tables S3 (inside) and
S4 (outside), converted from the VOC monitoring data using the paper's factor
of 1.9.

```{r table-3}
# `decimals` records how many decimal places the paper actually printed for
# each cell, transcribed from Table 3 by eye. It is carried explicitly rather
# than derived from CA_pub with format(), because format() is *vectorised* and
# pads every element to a common width -- which would silently give all twelve
# rows the same (wrong) precision.
table3_published <- tibble::tribble(
  ~date,        ~area,     ~air_ppm, ~CA_pub, ~decimals,
  "December 1", "Outside",     0.19,  0.0005,        4L,
  "December 2", "Outside",     3.04,  0.008,         3L,
  "December 3", "Outside",    57.00,  0.142,         3L,
  "December 4", "Outside",    12.92,  0.032,         3L,
  "December 5", "Outside",     1.33,  0.003,         3L,
  "December 6", "Outside",     0.19,  0.0005,        4L,
  "December 1", "Inside",      5.70,  0.014,         3L,
  "December 2", "Inside",     13.11,  0.033,         3L,
  "December 3", "Inside",     39.71,  0.1,           1L,
  "December 4", "Inside",   1649.20,  4.7,           1L,
  "December 5", "Inside",      0.19,  0.0005,        4L,
  "December 6", "Inside",      1.40,  0.003,         3L
)

table3 <- table3_published |>
  rowwise() |>
  mutate(
    CA_sim = peak_in_window(simulate_exposure(air_ppm, 24, tmax = 24), 24, "Cc"),
    # Round the simulation to the precision the paper printed, then compare.
    CA_sim_rounded = round(CA_sim, decimals),
    agrees = abs(CA_sim_rounded - CA_pub) < 1e-12
  ) |>
  ungroup()

table3 |>
  select(date, area, air_ppm, CA_pub, CA_sim, CA_sim_rounded, agrees) |>
  dplyr::rename(
    "Date" = date, "Area" = area, "VC (ppm)" = air_ppm,
    "CA published (mg/L)" = CA_pub, "CA simulated (mg/L)" = CA_sim,
    "CA simulated, rounded" = CA_sim_rounded,
    "Agrees at printed precision" = agrees
  ) |>
  knitr::kable(
    digits = 5,
    caption = "Table 3 of Boone et al. reproduced; 24 h continuous exposure."
  )
```

```{r table-3-assert, echo = FALSE}
stopifnot(nrow(table3) == 12L)
stopifnot(all(table3$agrees))
```

All twelve cells reproduce **exactly at the precision the paper printed**. The
Results narrative's two headline values also reproduce: the highest outside-area
prediction (57 ppm on December 3) and the highest inside-area prediction
(1649.2 ppm on December 4).

```{r table-3-headline}
table3 |>
  group_by(area) |>
  slice_max(CA_sim, n = 1) |>
  ungroup() |>
  transmute(
    Area = area, Date = date, `VC (ppm)` = air_ppm,
    `CA published (mg/L)` = CA_pub, `CA simulated (mg/L)` = CA_sim
  ) |>
  knitr::kable(digits = 4,
               caption = "Peak predicted arterial blood, inside vs outside the evacuation area.")
```

```{r table-3-headline-assert, echo = FALSE}
peaks <- table3 |> group_by(area) |> slice_max(CA_sim, n = 1) |> ungroup()
stopifnot(nrow(peaks) == 2L)
stopifnot(peaks$date[peaks$area == "Outside"] == "December 3")
stopifnot(peaks$date[peaks$area == "Inside"] == "December 4")
```

# Structural validation

PKNCA-style noncompartmental validation against published NCA is not available
for this paper - the authors report that they computed AUC to evaluate the
recoded model, but no AUC values appear in the paper or the supplement. The
checks below instead exercise the structural identities the Berkeley Madonna
listing itself carries as consistency checks.

## Mass balance

The listing defines `Mass = AF+AR+AS+AL+AM+AK+ASk+AVBlood` (chemical still in
tissues plus chemical already metabolised) and `InhDOSE' = QP*(CI-CX)` (net
amount absorbed), and states the identity `InhDose = Mass`. The packaged model
exposes these as `Amass` and `Ainhaled_net`.

```{r mass-balance}
mb <- simulate_exposure(1200, window_h = 1, tmax = 24)
mb_tail <- mb[nrow(mb), ]

tibble::tibble(
  Quantity = c("Amass (tissues + metabolised)", "Ainhaled_net (inhaled - exhaled)",
               "Relative closure error", "Cumulative metabolised (a_metabolized)"),
  Value = c(mb_tail$Amass, mb_tail$Ainhaled_net,
            (mb_tail$Amass - mb_tail$Ainhaled_net) / mb_tail$Ainhaled_net,
            mb_tail$a_metabolized),
  Units = c("mg", "mg", "-", "mg")
) |>
  knitr::kable(digits = c(0, 6, 0),
               caption = "Mass balance after a 1 h exposure at 1200 ppm, followed to 24 h.")
```

```{r mass-balance-assert, echo = FALSE}
stopifnot(abs((mb_tail$Amass - mb_tail$Ainhaled_net) / mb_tail$Ainhaled_net) < 1e-8)
stopifnot(mb_tail$Amass > 0)
```

## Blood-flow and tissue-volume balance

The listing carries `Qbal = Qtot - QC` and `Vbal = BW - Vtot` as checks that
the fractional flows and volumes partition cardiac output and body weight
exactly. Because the rapidly- and slowly-perfused groups are top-level splits
from which the named organs are carved out (`VS = VSC*BW - VF - VSk`,
`QR = QRC*QC - QL - QK`), these balances must close to machine precision.

```{r balance}
bal <- simulate_exposure(0, window_h = 1, tmax = 1)[1, ]

q_tot <- with(bal, q_slowly_perfused + q_rapidly_perfused + q_skin +
                q_liver + q_kidney + q_fat)
v_tot <- with(bal, v_slowly_perfused + v_rapidly_perfused + v_kidney +
                v_fat + v_skin + v_liver)

tibble::tibble(
  Check = c("Qtot - QC (L/h)", "BW - Vtot (L)"),
  Value = c(q_tot - bal$qc, 70 - v_tot)
) |>
  knitr::kable(digits = 10,
               caption = "The listing's Qbal and Vbal consistency checks.")
```

```{r balance-assert, echo = FALSE}
stopifnot(abs(q_tot - bal$qc) < 1e-9)
# Vbal does NOT close in the published listing: the fractional volumes sum to
# more than 1. See Errata; the residual is recorded rather than asserted away.
stopifnot(abs(70 - v_tot) > 1)
```

Blood flows partition cardiac output exactly. Tissue **volumes** do not
partition body weight - see Errata.

## Saturable metabolism

The Michaelis constant `km` is 0.04 mg/L. Sweeping air concentration over four
orders of magnitude shows where the liver crosses it, and therefore where the
model stops being linear in exposure.

```{r saturation}
sat <- tibble::tibble(air_ppm = c(1, 10, 100, 1000, 5000)) |>
  rowwise() |>
  mutate(
    .s = list(simulate_exposure(air_ppm, window_h = 1, tmax = 1)),
    CA = peak_in_window(.s, 1, "Cc"),
    CVL = peak_in_window(.s, 1, "cv_liver"),
    frac_of_vmax = max(.s$r_metab[.s$time < 1]) / .s$vmax[1]
  ) |>
  ungroup() |>
  mutate(CA_per_ppm = CA / air_ppm) |>
  select(air_ppm, CA, CA_per_ppm, CVL, frac_of_vmax)

sat |>
  dplyr::rename(
    "Air (ppm)" = air_ppm, "CA at 1 h (mg/L)" = CA,
    "CA per ppm" = CA_per_ppm,
    "Liver capillary conc (mg/L)" = CVL,
    "Metabolism as fraction of Vmax" = frac_of_vmax
  ) |>
  knitr::kable(digits = 7,
               caption = paste(
                 "Hepatic metabolism is first-order at community-level air",
                 "concentrations and saturated at AEGL-level concentrations."
               ))
```

```{r saturation-assert, echo = FALSE}
stopifnot(nrow(sat) == 5L)
# Below ~100 ppm the liver capillary concentration stays well under Km = 0.04,
# so metabolism runs at a small fraction of Vmax and the model is linear in
# exposure: CA per ppm is constant to better than 0.2%.
stopifnot(sat$CVL[sat$air_ppm == 100] < 0.04)
stopifnot(sat$frac_of_vmax[sat$air_ppm == 10] < 0.05)
lin <- sat$CA_per_ppm[sat$air_ppm %in% c(1, 10, 100)]
stopifnot(abs(max(lin) / min(lin) - 1) < 0.002)
# At AEGL-level concentrations the liver is saturated ...
stopifnot(sat$frac_of_vmax[sat$air_ppm == 1000] > 0.97)
stopifnot(sat$CVL[sat$air_ppm == 1000] > 0.04)
# ... and arterial blood becomes supra-proportional: CA per ppm rises
# monotonically, and by more than 10% from 1 ppm to 5000 ppm.
stopifnot(all(diff(sat$CA_per_ppm) > 0))
stopifnot(sat$CA_per_ppm[sat$air_ppm == 5000] /
            sat$CA_per_ppm[sat$air_ppm == 1] > 1.10)
```

Below about 100 ppm the liver capillary concentration stays well under `km`,
metabolism runs at a few percent of `vmax`, and arterial blood is proportional
to air concentration to better than 0.2%. At AEGL-level concentrations the
liver is saturated (over 97% of `vmax` at 1000 ppm) and arterial blood becomes
supra-proportional.

This is visible in the published Table 3 itself. On the low-concentration days
the predicted blood level tracks air concentration linearly -- 3.04 ppm gives
0.008 mg/L and 57 ppm gives 0.142 mg/L, both about 0.0025 mg/L per ppm -- while
the saturating December 4 inside-area value of 1649.2 ppm gives 4.7 mg/L, about
0.0029 mg/L per ppm.

```{r saturation-published-assert, echo = FALSE}
ratio <- function(area, date) {
  r <- table3[table3$area == area & table3$date == date, ]
  stopifnot(nrow(r) == 1L)
  r$CA_pub / r$air_ppm
}
low <- c(ratio("Outside", "December 2"), ratio("Outside", "December 3"))
high <- ratio("Inside", "December 4")
stopifnot(abs(low[1] / low[2] - 1) < 0.07)  # the two low-level days agree
stopifnot(high > max(low) * 1.08)           # the saturating day does not
```

## Exhaled breath as a biomarker

The paper proposes exhaled breath as a noninvasive exposure biomarker. The
listing's sampled-breath equation mixes 70% alveolar air with 30% unabsorbed
inhaled air and reports the result in ppm (`CXppm`); the packaged model exposes
this as `Cexhaled_ppm`. During exposure it is dominated by the inhaled air
itself, and only after exposure stops does it report the body burden.

```{r breath}
br <- simulate_exposure(1200, window_h = 1, tmax = 4)

ggplot(br, aes(time, Cexhaled_ppm)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Sampled exhaled breath (ppm)",
    title = "Exhaled-breath VC after a 1 h exposure at 1200 ppm",
    caption = "Dashed line marks the end of the exposure window."
  ) +
  theme_minimal()
```

```{r breath-assert, echo = FALSE}
# During exposure the sampled breath is dominated by the 30% inhaled-air term,
# so it must exceed 30% of the inhaled 1200 ppm.
stopifnot(max(br$Cexhaled_ppm[br$time < 1]) > 0.3 * 1200)
# After exposure it falls by more than an order of magnitude within 3 h.
post_peak <- br$Cexhaled_ppm[br$time >= 1][1]
stopifnot(br$Cexhaled_ppm[nrow(br)] < post_peak / 10)
```

## Exposure metrics via PKNCA

For completeness, standard exposure metrics are computed with PKNCA on the
arterial-blood profile for the three 60-minute AEGL scenarios. There is no
published counterpart to compare against; these characterise the packaged model
rather than validate it.

```{r pknca}
pknca_sims <- tibble::tribble(
  ~scenario,  ~air_ppm,
  "AEGL-1",     250,
  "AEGL-2",    1200,
  "AEGL-3",    4800
) |>
  rowwise() |>
  reframe(
    scenario = scenario,
    simulate_exposure(air_ppm, window_h = 1, tmax = 24, dt = 0.02) |>
      select(time, Cc)
  ) |>
  filter(!is.na(Cc)) |>
  mutate(id = 1L)

conc_obj <- PKNCA::PKNCAconc(pknca_sims, Cc ~ time | scenario + id)

intervals <- data.frame(
  start = 0, end = 24,
  cmax = TRUE, tmax = TRUE, auclast = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, intervals = intervals))

as.data.frame(nca_res) |>
  select(scenario, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::rename(
    "Scenario" = scenario, "Cmax (mg/L)" = cmax, "Tmax (h)" = tmax,
    "AUC0-24 (mg*h/L)" = auclast
  ) |>
  knitr::kable(digits = 3,
               caption = "Arterial-blood exposure metrics, 60 min AEGL scenarios.")
```

```{r pknca-assert, echo = FALSE}
nca_df <- as.data.frame(nca_res)
stopifnot(nrow(nca_df) == 9L)
cmax_vals <- nca_df$PPORRES[nca_df$PPTESTCD == "cmax"]
stopifnot(length(cmax_vals) == 3L, all(!is.na(cmax_vals)), all(cmax_vals > 0))
# Cmax must be ordered AEGL-1 < AEGL-2 < AEGL-3.
stopifnot(!is.unsorted(cmax_vals[order(c(250, 1200, 4800))]))
```

# Assumptions and deviations

* **Exposure enters as a parameter, not a dose record.** The published model is
  driven by an air concentration and window length, so the vignette's event
  tables contain observation rows only and every scenario is set through
  `rxSolve(params = )`. No `evid = 1` record appears anywhere.
* **Body weight.** The listing hard-codes `BW = 70.0`; every simulation here
  uses `WT = 70` to match. `WT` is carried as a covariate so the model can be
  re-scaled, but no published simulation exercises a different weight, so the
  weight-scaling path is untested against the paper.
* **The oral and dermal routes are never exercised by the paper.** They are
  implemented faithfully from the listing (`r_oral`, `r_dermal`,
  `r_dermal_out`, and the `a_oral` / `a_dermal_absorbed` /
  `a_dermal_eliminated` accumulators) but every published simulation switches
  them off. Their correctness rests on transcription from the listing alone,
  not on any reproduced published output.
* **No between-subject variability and no residual error.** The model is a
  recoded deterministic literature model, evaluated by the authors against
  digitised human kinetic data rather than fitted. The paper handles
  interindividual toxicokinetic variability outside the model with an
  assessment factor of 3.16. Nothing was invented to fill the gap.
* **Output grid.** The Berkeley Madonna listing used `dtout = 0.01` h with
  `METHOD RK4`; this vignette uses a 0.01 h output grid with rxode2's default
  solver. Residual differences at the shortest exposure window are consistent
  with this.

# Errata and observations on the source

* **Table 4, 4-hour rows.** All three report `CA < CV`, which is
  thermodynamically impossible during uptake. As demonstrated above, CA and CX
  in those rows were read one output step *after* the exposure window closed
  while CV was read at the window end; under that readout the published values
  reproduce to better than 1.2%. The rest of Table 4 uses the
  peak-during-exposure readout. This is a reporting inconsistency in the paper,
  not a defect in the model.
* **Table 4, 10-minute rows.** CA and CX reproduce within 2.3%, but the
  published CV values sit 6.7-6.9% above the simulation across all three AEGL
  tiers - a consistent offset that suggests a readout detail at the shortest
  window that the listing does not record. The same model reproduces the
  30-minute, 60-minute and 8-hour rows to better than 0.5%, so this is
  localised to the shortest window rather than a structural problem.
* **Table 4, AEGL-1 4 h and 8 h.** The published CA rises from 0.11 (4 h) to
  0.17 (8 h) even though the air level *falls* from 140 to 70 ppm. This is a
  consequence of the readout inconsistency above rather than a modelling claim:
  the 4-hour value is a post-exposure reading and the 8-hour value is an
  in-exposure reading.
* **Tissue volumes do not partition body weight.** The listing's fractional
  volumes sum to well over 1 (`VSC = 0.82` plus `VRC = 0.09` plus
  `VBloodC = 0.079` already exceeds 1 before fat, skin, liver and kidney are
  carved out), so the listing's own `Vbal = BW - Vtot` check does not close.
  The carve-out structure (`VS = VSC*BW - VF - VSk`) keeps every individual
  volume positive and the kinetics well-posed, and the blood-flow balance
  `Qbal` does close exactly. The volumes are transcribed exactly as published.
* **Kidney volume fraction.** `VKC = 0.044` is roughly an order of magnitude
  above the human kidney fraction in Brown et al. 1997 (about 0.0044), the
  source the listing names for its physiological parameters. It is transcribed
  as published; it inflates `v_kidney` but has little effect on arterial blood
  because the kidney is a small, non-eliminating compartment in this structure.
* **Metabolite kinetics are deliberately absent.** Methods 2.2 states that
  "the original model simulations for metabolites and metabolite data were not
  included in the generic PBPK model", so `a_metabolized` accumulates the
  parent consumed by hepatic metabolism without any downstream metabolite
  disposition. The reactive metabolites of VC drive its carcinogenicity, so
  this model must not be used for metabolite-based risk metrics.
* **Vmax units.** The listing annotates `Vmaxc = 3.97` as "mg/h" but uses it as
  `VMAX = Vmaxc*BW**.75`, i.e. it is an allometric coefficient in mg/h/kg^0.75.
  The label in the model file records the operative units.
