---
title: "Tigecycline (Borsuk-De Moor 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tigecycline (Borsuk-De Moor 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
rxode2::rxSetSeed(20180422)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("BorsukDeMoor_2018_tigecycline"))$reference`
* Description: `r rxode2::rxode(readModelDb("BorsukDeMoor_2018_tigecycline"))$description`
* Article: <https://doi.org/10.1128/AAC.02273-17>

This is the third tigecycline model in the library fitted to a critically ill
cohort, and the only one of the four fitted to the *high-dose* regimen. The
licensed regimen is a 100 mg loading dose followed by 50 mg every 12 h;
`Luo_2023_tigecycline`, `Song_2024_tigecycline` and `Su_2024_tigecycline` were
all fitted on that schedule. Borsuk-De Moor 2018 doubles it throughout - 200 mg
loading, then 100 mg every 12 h - following the Ramirez 2013 and De Pascale 2014
findings that the licensed dose underexposes ICU patients.

Two features distinguish it structurally from its siblings:

* **Interoccasion variability.** Samples were drawn after *every* dose, so the
  authors could split within-patient from between-patient variability. Doing so
  moved a substantial share of clearance variability out of IIV: the remaining
  IIV on CL is 17.3%, roughly half of what comparable studies report, with a
  further 14.4% carried as IOV.
* **No retained covariate.** Twelve covariates were screened - including the
  weight, creatinine clearance and sex terms that Van Wart 2006 and Rubino 2010
  retained - and none showed a systematic relationship. The paper's conclusion
  is that uniform dosing is sufficient in this population and that adjustment
  should be driven by the pathogen MIC instead.

# Population

Thirty-seven adults with sepsis or septic shock in two Polish tertiary
medical/surgical ICUs (Lublin and Olsztyn), contributing 940 analysed plasma
tigecycline concentrations from 942 collected (two removed as CWRES > 5
outliers). Median age 61 years (range 25-79), weight 80 kg (50-129), height
175 cm (158-190); 11 of 37 (29.7%) were female. The cohort was severely ill:
median SOFA score 13 points (range 2.0-21), median albumin 2.2 g/dL (1.5-3.6),
median procalcitonin 8.22 ug/L (0.16-122), and 23 of 37 patients died. Thirty of
37 were on continuous renal replacement therapy and 2 on ECMO.

Every patient received the same high-dose regimen: a 200 mg loading dose as a
30-minute intravenous infusion, then 100 mg every 12 h as 30-minute infusions,
for 2 to 8 doses over 1 to 4 consecutive days. Arterial blood was sampled at
0.5, 2, 4, 8 and 12 h after *every* dose - the dense per-occasion schedule that
makes the IOV estimable.

The same information is available programmatically via
`readModelDb("BorsukDeMoor_2018_tigecycline")()$population`.

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/BorsukDeMoor_2018_tigecycline.R` carries an in-file
comment naming its origin. They are collected here for review. All values come
from the main article; the supplemental material (`AAC.02273-17_zac004187006s1.pdf`)
contains only the chromatographic method, its validation, and Figures S1-S6, and
carries no NONMEM control stream and no additional parameter values.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Two-compartment linear disposition, clearance/volume parameterisation | `ADVAN3 TRANS4` | Materials and Methods, *Pharmacokinetic modeling*: "a two-compartment model was used"; "parametrized in terms of clearances (CL and Q) and volumes of distribution (V1 and V2)" |
| `lcl` = log(CL) | 22.1 L/h | Table 2, `theta CL` (RSE 3.16%; bootstrap 22.1, 90% CI 20.9-23.2) |
| `lvc` = log(V1) | 162 L | Table 2, `theta V1` (RSE 5.3%; bootstrap 163, 90% CI 150-176) |
| `lq` = log(Q) | 69.4 L/h | Table 2, `theta Q` (RSE 32.6%; bootstrap 67.3, 90% CI 41.9-98.4) |
| `lvp` = log(V2) | 87.9 L | Table 2, `theta V2` (RSE 8.67%; bootstrap 87.6, 90% CI 76.1-101) |
| IIV form `P = theta * exp(eta) * exp(kappa)` | log-normal | Materials and Methods: "P(i,k) = theta_P exp(eta_P,i) exp(kappa_P,i,k)" |
| `etalcl` | 17.3 % CV -> 0.029929 | Table 2, `omega^2 CL (% CV)` (RSE 19%, shrinkage 7.3%) |
| `etalvc` | 19.2 % CV -> 0.036864 | Table 2, `omega^2 V1 (% CV)` (RSE 29.2%, shrinkage 6.7%) |
| `etalvp` | 38.7 % CV -> 0.149769 | Table 2, `omega^2 V2 (% CV)` (RSE 40.8%, shrinkage 22.4%) |
| No IIV on Q | fixed at 0 | Table 2, `omega^2 Q (% CV)` = "0 FIX"; Results: "it was not possible to estimate the interindividual variability (IIV) for the intercompartmental clearance (Q2)" |
| `etaiov_lcl_1` .. `etaiov_lcl_8` | 14.4 % CV -> 0.020736 | Table 2, `pi^2 CL (% CV)` (RSE 35%), with per-occasion shrinkage printed for Occasions 1-8 |
| `etaiov_lvp_1` .. `etaiov_lvp_8` | 20.8 % CV -> 0.043264 | Table 2, `pi^2 V2 (% CV)` (RSE 66.4%), with per-occasion shrinkage printed for Occasions 1-8 |
| Occasion count = 8; shared magnitude across occasions | - | Table 2 prints Occasion 1-8 shrinkage rows; Materials and Methods: "The IIV and IOV variances were assumed to be constant across occasions" |
| `addSd` | 0.0210 mg/L | Table 2, `sigma_add (ug/ml)`; footnote a: "sigma_add, additive residual random error" |
| `propSd` | 13.0 % CV -> 0.130 | Table 2, `sigma^2_prop (% CV)`; footnote a: "sigma^2_prop, variance of proportional residual random error" |
| Combined additive + proportional residual | - | Materials and Methods: "modeled using a combined additional and proportional error model" |
| Dosing regimen simulated below | 200 mg then 7 x 100 mg q12h, 30-min infusions | Patients and study design; Figure 6 caption |
| Published `Vss` | 250 L | Discussion: "Mean Vss in our study is 250 liters" |
| Published accumulation ratio | 1.57 | Discussion: "accumulation ratio of 1.57 in our study compared to 4.00 [18] and 2.55 [19]" |

# Virtual cohort

The cohort reproduces the study's own regimen and its eight dosing occasions.
No covariate enters the model, so the only subject-level column needed is the
occasion indicator `OCC`, which is time-varying: occasion `k` spans
`[(k-1) * 12, k * 12)` hours, matching the paper's definition of an occasion as
one 12-hourly administration.

```{r cohort}
n_sub <- 150L
tau <- 12      # dosing interval (h)
t_inf <- 0.5   # infusion duration (h)
load_mg <- 200
maint_mg <- 100
n_doses <- 8L  # loading dose + 7 maintenance doses, per Figure 6
t_end <- as.numeric(n_doses) * tau

dose_times <- seq(0, by = tau, length.out = n_doses)
dose_amts <- c(load_mg, rep(maint_mg, n_doses - 1L))

# Occasion index for an arbitrary time, clamped to the eight modelled slots.
occ_of <- function(time) pmin(pmax(floor(time / tau) + 1L, 1L), 8L)

subj <- data.frame(id = seq_len(n_sub))

doses <- subj |>
  tidyr::crossing(dose_idx = seq_len(n_doses)) |>
  dplyr::mutate(
    time = dose_times[dose_idx],
    amt = dose_amts[dose_idx],
    evid = 1L,
    # Explicit infusion rate (mg/h). An IV dose given without a rate is
    # delivered as a bolus, which every AUC-based gate would still pass, so
    # the rate is set here rather than left to a duration column.
    rate = amt / t_inf,
    cmt = "central"
  ) |>
  dplyr::select(-dose_idx)

obs_times <- seq(0, t_end, by = 0.5)
obs <- subj |>
  tidyr::crossing(time = obs_times) |>
  dplyr::mutate(amt = NA_real_, evid = 0L, rate = NA_real_, cmt = "central")

events <- dplyr::bind_rows(doses, obs) |>
  dplyr::mutate(OCC = occ_of(time)) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

# All eight occasion slots must actually be exercised, otherwise the IOV
# structure is only nominally tested.
stopifnot(
  setequal(unique(events$OCC), 1:8),
  dplyr::n_distinct(events$id) == n_sub,
  all(events$rate[events$evid == 1] == events$amt[events$evid == 1] / t_inf)
)
```

# Simulation

```{r simulate}
# `readModelDb()` returns the model FUNCTION; `rxode()` turns it into the rxUi
# that carries `$simulationModel`, `$omega` and `$theta`.
mod <- rxode2::rxode(readModelDb("BorsukDeMoor_2018_tigecycline"))
stopifnot(inherits(mod, "rxUi"), !is.null(mod$simulationModel))
# 3 IIV etas + 2 parameters x 8 occasion slots of IOV.
stopifnot(identical(dim(mod$omega), c(19L, 19L)))

# `omega` is passed explicitly: rxode2 caches the previous solve's omega against
# the compiled model, so a later population solve can silently collapse onto one
# typical subject (and a later typical-value solve can silently re-sample etas).
sim <- rxode2::rxSolve(
  mod$simulationModel, events,
  omega = mod$omega, keep = "OCC"
) |>
  as.data.frame()

stopifnot(all(c("Cc", "sim", "cl", "vc", "q", "vp", "OCC") %in% names(sim)))
# Cc is exactly 0 at t = 0: the observation sits at the start of the first
# 30-minute infusion, before any drug has been delivered.
stopifnot(!anyNA(sim$Cc), all(sim$Cc >= 0))
stopifnot(all(sim$Cc[sim$time > 0] > 0))
stopifnot(all(sim$Cc[sim$time == 0] == 0))

# Guard the population direction of the omega-caching bug: IIV must have varied.
stopifnot(dplyr::n_distinct(round(sim$vc, 8)) > 1L)
# Guard the IOV direction: clearance must vary WITHIN a subject across
# occasions, which only IOV can produce (no covariate acts on CL).
within_subject_cl_levels <- sim |>
  dplyr::group_by(id) |>
  dplyr::summarise(n_cl = dplyr::n_distinct(round(cl, 8)), .groups = "drop")
stopifnot(all(within_subject_cl_levels$n_cl == 8L))
```

# Replicate published figures

## Figure 6 - typical-value concentration-time profile

Figure 6 of Borsuk-De Moor 2018 simulates tigecycline concentrations after the
200 mg loading dose and seven subsequent 100 mg doses at 12-hour intervals,
using the typical population parameters (the black solid line of that figure;
the red dashed lines are the three comparator models, which are not reproduced
here).

```{r figure-6}
mod_typ <- rxode2::zeroRe(mod)
typ_events <- events |> dplyr::filter(id == 1L)

# `omega = NA` is the only reliable way to force a typical-value solve;
# zeroRe() alone is not sufficient once a stochastic solve has run.
typ <- rxode2::rxSolve(mod_typ, typ_events, omega = NA, keep = "OCC") |>
  as.data.frame()

# If etas were silently re-sampled, cl would vary across occasions here too.
stopifnot(dplyr::n_distinct(round(typ$cl, 8)) == 1L)
stopifnot(dplyr::n_distinct(round(typ$vp, 8)) == 1L)

ggplot(typ, aes(time, Cc)) +
  geom_line() +
  geom_vline(xintercept = dose_times, linetype = "dotted", alpha = 0.4) +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Tigecycline (mg/L)",
    title = "Typical-value profile, 200 mg then 7 x 100 mg q12h",
    caption = "Replicates the black solid line of Figure 6 of Borsuk-De Moor 2018."
  )
```

## Figure 2 - visual predictive check

Figure 2 is a VPC over a dosing interval. The observed concentrations behind it
are not public, so only the simulated side is reproduced: the 5th, 50th and 95th
percentiles of simulated observations (`sim`, the individual prediction plus a
combined additive-and-proportional residual draw).

```{r figure-2}
sim |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    Q05 = quantile(sim, 0.05),
    Q50 = quantile(sim, 0.50),
    Q95 = quantile(sim, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Tigecycline (mg/L)",
    title = "Simulated 5th / 50th / 95th percentiles over the 96 h course",
    caption = "Replicates the predicted percentiles of Figure 2 of Borsuk-De Moor 2018."
  )
```

# Structural validation

## Closed-form biexponential

rxode2 recognises a `cl` / `vc` / `q` / `vp` parameter set and replaces the
written ODE system with its analytic solution. That substitution is silent, and
when a `q` / `vp` pair is *missing* it collapses a two-compartment model to one
compartment while leaving every AUC-based gate passing. The check below is
therefore the load-bearing structural test: the typical-value solve is compared
against the two-compartment infusion biexponential evaluated independently from
the `ini()` values.

Both sides use the same parameters, so the only difference is numerical
integration error and the bound is tight.

```{r closed-form}
theta <- mod$theta
cl_t <- exp(theta[["lcl"]])
vc_t <- exp(theta[["lvc"]])
q_t <- exp(theta[["lq"]])
vp_t <- exp(theta[["lvp"]])

k10 <- cl_t / vc_t
k12 <- q_t / vc_t
k21 <- q_t / vp_t

# Hybrid rate constants of the two-compartment system.
a_sum <- k10 + k12 + k21
b_prod <- k10 * k21
lam1 <- (a_sum + sqrt(a_sum^2 - 4 * b_prod)) / 2   # distribution
lam2 <- (a_sum - sqrt(a_sum^2 - 4 * b_prod)) / 2   # terminal
stopifnot(lam1 > lam2, lam2 > 0)

coefA <- (lam1 - k21) / (lam1 - lam2)
coefB <- (k21 - lam2) / (lam1 - lam2)

# Superposition of constant-rate infusions: contribution of one dose of `amt`
# infused over `t_inf` at `t0`, evaluated at absolute time `tt`.
inf_conc <- function(tt, t0, amt) {
  r0 <- amt / t_inf
  te <- pmax(tt - t0, 0)                 # time since infusion start
  tin <- pmin(te, t_inf)                 # time spent infusing so far
  toff <- pmax(te - t_inf, 0)            # time since infusion end
  (r0 / vc_t) * (
    (coefA / lam1) * (1 - exp(-lam1 * tin)) * exp(-lam1 * toff) +
      (coefB / lam2) * (1 - exp(-lam2 * tin)) * exp(-lam2 * toff)
  )
}

cf <- rowSums(vapply(
  seq_len(n_doses),
  function(i) inf_conc(typ$time, dose_times[i], dose_amts[i]),
  numeric(nrow(typ))
))

# Drop t = 0, where both sides are exactly 0 and a relative error is undefined.
keep <- typ$time > 0
stopifnot(sum(keep) > 100, all(cf[keep] > 0))
rel_err <- abs(typ$Cc[keep] - cf[keep]) / cf[keep]
stopifnot(all(is.finite(rel_err)), max(rel_err) < 1e-5)
# t = 0 must still agree exactly.
stopifnot(all(abs(typ$Cc[!keep] - cf[!keep]) < 1e-12))
```

```{r closed-form-report}
data.frame(
  Quantity = c(
    "Distribution rate constant (1/h)", "Terminal rate constant (1/h)",
    "Distribution half-life (h)", "Terminal half-life (h)",
    "Max relative difference, solve vs closed form"
  ),
  Value = c(
    lam1, lam2, log(2) / lam1, log(2) / lam2, max(rel_err)
  )
) |>
  knitr::kable(
    digits = c(0, 6),
    caption = "Two-compartment hybrid constants and the closed-form agreement bound."
  )
```

## Steady-state volume of distribution

The Discussion states `Vss` = 250 L, defined in Materials and Methods as the sum
of `V1` and `V2`. This is the only volume quantity the paper reports beyond the
two Table 2 estimates, and it confirms both are read on the same scale.

```{r vss}
vss <- vc_t + vp_t
stopifnot(abs(vss - 250) < 0.5)
```

The model gives `Vss` = `r sprintf("%.1f", vss)` L against the published 250 L.

## Accumulation ratio

The Discussion reports an accumulation ratio of 1.57 on this regimen, contrasted
with 4.00 for Van Wart 2006 and 2.55 for Rubino 2010, and attributes the
difference to this study's much smaller peripheral volume. The ratio follows
from the terminal rate constant and the dosing interval,
`R = 1 / (1 - exp(-lambda_z * tau))`.

This is the strongest published gate available, because `lambda_z` is a joint
function of all four structural parameters: mis-transcribing any one of `CL`,
`V1`, `Q` or `V2` moves it. `Vss` above, by contrast, would survive a swap of
`V1` and `V2`.

```{r accumulation}
accum <- 1 / (1 - exp(-lam2 * tau))
stopifnot(abs(accum - 1.57) < 0.01)
```

The model gives an accumulation ratio of `r sprintf("%.3f", accum)` against the
published 1.57.

## Simulated accumulation

The analytic ratio above is a terminal-phase identity. The simulated profile
gives the same quantity empirically, as the ratio of `AUC` over the last dosing
interval to `AUC` over the first, for a 100 mg dose on both. This needs a
maintenance-only schedule, so the loading dose is dropped for this check alone.

```{r accumulation-sim}
maint_doses <- subj |>
  dplyr::filter(id == 1L) |>
  tidyr::crossing(time = dose_times) |>
  dplyr::mutate(
    amt = maint_mg, evid = 1L, rate = maint_mg / t_inf, cmt = "central"
  )
maint_obs <- subj |>
  dplyr::filter(id == 1L) |>
  tidyr::crossing(time = seq(0, t_end, by = 0.05)) |>
  dplyr::mutate(amt = NA_real_, evid = 0L, rate = NA_real_, cmt = "central")
maint_ev <- dplyr::bind_rows(maint_doses, maint_obs) |>
  dplyr::mutate(OCC = occ_of(time)) |>
  dplyr::arrange(time, dplyr::desc(evid)) |>
  as.data.frame()

maint <- rxode2::rxSolve(mod_typ, maint_ev, omega = NA) |> as.data.frame()

auc_window <- function(d, lo, hi) {
  w <- d[d$time >= lo & d$time <= hi, ]
  sum(diff(w$time) * (utils::head(w$Cc, -1) + utils::tail(w$Cc, -1)) / 2)
}
auc_first <- auc_window(maint, 0, tau)
auc_last <- auc_window(maint, t_end - tau, t_end)
accum_sim <- auc_last / auc_first

# Approaches but has not fully reached the terminal-phase asymptote after eight
# intervals, so it sits just below the analytic value.
stopifnot(accum_sim > 1.50, accum_sim < accum)
```

Simulated AUC accumulation over eight intervals is
`r sprintf("%.3f", accum_sim)`, against the analytic asymptote of
`r sprintf("%.3f", accum)` and the published 1.57.

# PKNCA validation

NCA is run on the typical-value single-dose profile, which is where the paper's
two published derived quantities (`CL` and `Vss`) are directly recoverable, and
on the stochastic cohort, to give the steady-state exposure summary a user of
this model will want.

```{r pknca-single}
single_ev <- data.frame(
  id = 1L,
  time = c(0, seq(0, 240, by = 0.05)),
  amt = c(load_mg, rep(NA_real_, length(seq(0, 240, by = 0.05)))),
  evid = c(1L, rep(0L, length(seq(0, 240, by = 0.05)))),
  rate = c(load_mg / t_inf, rep(NA_real_, length(seq(0, 240, by = 0.05)))),
  cmt = "central",
  OCC = 1L
)
single <- rxode2::rxSolve(mod_typ, single_ev, omega = NA) |> as.data.frame()

# A one-subject solve returns no `id` column, so PKNCA's grouping key has to be
# added back explicitly.
stopifnot(!"id" %in% names(single))
single$id <- 1L

sc <- single |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc) |>
  dplyr::distinct(id, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time) |>
  as.data.frame()

conc_single <- PKNCA::PKNCAconc(sc, Cc ~ time | id)
dose_single <- PKNCA::PKNCAdose(
  data.frame(id = 1L, time = 0, amt = load_mg, duration = t_inf),
  amt ~ time | id, duration = "duration", route = "intravascular"
)

# The terminal phase of a two-compartment profile is only clean well after
# distribution is complete; an automatic lambda-z window would otherwise wander
# into the distribution phase and bias half-life (and therefore Vss).
iv_intervals <- data.frame(
  start = 0, end = Inf,
  aucinf.obs = TRUE, cl.obs = TRUE, vss.obs = TRUE,
  half.life = TRUE, mrt.iv.obs = TRUE,
  impute = NA_character_
)
nca_single <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  conc_single, dose_single, intervals = iv_intervals
))
single_tbl <- as.data.frame(nca_single$result)
```

```{r pknca-single-gate}
getp <- function(code) {
  v <- single_tbl$PPORRES[single_tbl$PPTESTCD == code]
  stopifnot(length(v) == 1L, is.finite(v))
  v
}

# Mass balance: for an IV dose, CL * AUCinf must return the dose exactly.
# This is a pure closed-form identity for a linear model, so the bound is tight.
stopifnot(abs(cl_t * getp("aucinf.obs") / load_mg - 1) < 1e-3)
# NCA half-life must recover the analytic terminal half-life.
stopifnot(abs(getp("half.life") - log(2) / lam2) < 0.05)
```

## Comparison against the published derived parameters

Borsuk-De Moor 2018 publishes no NCA table - no Cmax, Tmax, AUC or half-life is
reported for the cohort - so the reference side below is the paper's two
published derived quantities: the Table 2 clearance estimate and the Discussion's
`Vss`. Recovering both from an independent NCA of the simulated profile confirms
that the model as encoded reproduces the paper's own summary of its fit.

One correction is needed first. PKNCA's `vss.obs` is defined as
`cl.obs * mrt.obs`, which for an infusion does *not* subtract the
infusion-duration term; on this 30-minute infusion it overstates `Vss` by
`CL * t_inf / 2` = 5.5 L, or 2.2%. The steady-state volume the paper reports is
`V1 + V2`, the infusion-corrected quantity, so the comparison below uses
`cl.obs * mrt.iv.obs` instead.

```{r nca-comparison}
reference_nca <- data.frame(
  cl.obs = 22.1,   # Borsuk-De Moor 2018 Table 2, theta CL
  vss.obs = 250    # Borsuk-De Moor 2018 Discussion, "Mean Vss in our study is 250 liters"
)

# PKNCA's `vss.obs` is `cl.obs * mrt.obs`, WITHOUT the infusion-duration
# correction - assert that definition rather than assume it.
stopifnot(abs(getp("vss.obs") - getp("cl.obs") * getp("mrt.obs")) < 1e-6)

# For a 30-minute infusion the uncorrected form overstates Vss by
# CL * t_inf / 2 = 5.5 L. The paper's Vss is V1 + V2, the corrected quantity,
# so the comparison uses `mrt.iv.obs`, which PKNCA reports separately and which
# does carry the correction.
simulated_nca <- data.frame(
  cl.obs = getp("cl.obs"),
  vss.obs = getp("cl.obs") * getp("mrt.iv.obs")
)

nca_cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = simulated_nca,
  reference = reference_nca,
  params = c("cl.obs", "vss.obs"),
  units = c(cl.obs = "L/h", vss.obs = "L")
)
knitr::kable(
  nca_cmp,
  digits = 2,
  caption = paste(
    "Simulated NCA of the typical-value 200 mg profile against the derived",
    "parameters published by Borsuk-De Moor 2018."
  )
)
```

```{r nca-comparison-gate}
# `% diff` is a formatted DISPLAY string ("+0.0%", with a trailing "*" when the
# tolerance is exceeded), so it must be parsed rather than coerced: as.numeric()
# on it returns NA, and `all(NA < 1)` is NA, which would fail loudly here but
# pass vacuously under any assertion wrapped in isTRUE() or any().
pct <- as.numeric(sub("%$", "", sub("\\*$", "", nca_cmp[["% diff"]])))
stopifnot(length(pct) == 2L, !anyNA(pct))

# Both published quantities must be recovered, not merely be close.
stopifnot(all(abs(pct) < 1))
# No row may carry the >20% tolerance flag.
stopifnot(!any(grepl("*", nca_cmp[["% diff"]], fixed = TRUE)))
```

```{r nca-comparison-footnote, results = "asis"}
fn <- attr(nca_cmp, "footnote")
if (!is.null(fn)) cat(fn, "\n")
```

## Steady-state exposure across the cohort

```{r pknca-ss}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc), time >= t_end - tau) |>
  dplyr::mutate(time = time - (t_end - tau)) |>
  dplyr::select(id, time, Cc) |>
  dplyr::distinct(id, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time) |>
  as.data.frame()

# A time = 0 row is present by construction here (the grid lands on 84 h);
# the defensive bind is a no-op rather than an imputation.
stopifnot(all(sim_nca$time[!duplicated(sim_nca$id)] == 0))

conc_ss <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | id)
dose_ss <- PKNCA::PKNCAdose(
  data.frame(id = seq_len(n_sub), time = 0, amt = maint_mg, duration = t_inf),
  amt ~ time | id, duration = "duration", route = "intravascular"
)
ss_intervals <- data.frame(
  start = 0, end = tau,
  cmax = TRUE, cmin = TRUE, cav = TRUE, tmax = TRUE, auclast = TRUE
)
nca_ss <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  conc_ss, dose_ss, intervals = ss_intervals
))
ss_tbl <- as.data.frame(nca_ss$result)
```

```{r pknca-ss-summary}
ss_tbl |>
  dplyr::filter(PPTESTCD %in% c("cmax", "cmin", "cav", "tmax", "auclast")) |>
  dplyr::group_by(PPTESTCD) |>
  dplyr::summarise(
    median = median(PPORRES),
    p05 = quantile(PPORRES, 0.05),
    p95 = quantile(PPORRES, 0.95),
    .groups = "drop"
  ) |>
  dplyr::rename(
    "Parameter" = PPTESTCD, "Median" = median,
    "5th pctile" = p05, "95th pctile" = p95
  ) |>
  knitr::kable(
    digits = 2,
    caption = paste(
      "Simulated steady-state NCA over the eighth 12 h dosing interval, n = 150.",
      "Cmax / Cmin / Cav in mg/L, AUC0-12 in mg*h/L, Tmax in h.",
      "Borsuk-De Moor 2018 publishes no NCA table to compare against."
    )
  )
```

```{r pknca-ss-gate}
cav <- ss_tbl$PPORRES[ss_tbl$PPTESTCD == "cav"]
auc <- ss_tbl$PPORRES[ss_tbl$PPTESTCD == "auclast"]
stopifnot(length(cav) == n_sub, !anyNA(cav))

# At steady state on a linear model, AUC0-tau = Dose / CL. Assert on the CENTRE
# of the cohort rather than on its extremes: the per-subject ratio varies with
# each subject's own drawn clearance, and the extreme of a random cohort is not
# reproducible across rxode2 builds.
ratio <- auc * (cl_t / maint_mg)
stopifnot(abs(median(ratio) - 1) < 0.10)
stopifnot(quantile(abs(ratio - 1), 0.9) < 0.40)
```

# Interoccasion variability

IOV is what separates this model from its three tigecycline siblings, so it is
worth showing that it is actually active. For each subject, clearance takes a
different value on each of the eight occasions while the subject's own `eta`
stays fixed; the spread *within* a subject is the IOV, and its magnitude should
match the 14.4% CV of Table 2.

```{r iov}
iov_cv <- sim |>
  dplyr::group_by(id, OCC) |>
  dplyr::summarise(cl = dplyr::first(cl), .groups = "drop") |>
  dplyr::group_by(id) |>
  # Within-subject SD of log(CL) across occasions estimates the IOV omega.
  dplyr::summarise(sd_log_cl = sd(log(cl)), .groups = "drop")

# The within-subject SD of log(CL) across 8 occasions has a wide sampling
# distribution at n = 8, so assert on the cohort median against the published
# 0.144, not on any subject's own value.
stopifnot(abs(median(iov_cv$sd_log_cl) - 0.144) < 0.04)
```

```{r iov-plot}
sim |>
  dplyr::filter(id <= 12L) |>
  dplyr::group_by(id, OCC) |>
  dplyr::summarise(cl = dplyr::first(cl), .groups = "drop") |>
  ggplot(aes(factor(OCC), cl, group = id)) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1) +
  geom_hline(yintercept = cl_t, linetype = "dashed") +
  labs(
    x = "Occasion", y = "Clearance (L/h)",
    title = "Interoccasion variability in clearance, first 12 subjects",
    caption = paste(
      "Each line is one subject across the eight dosing occasions.",
      "Dashed line is the typical value, 22.1 L/h."
    )
  )
```

Median within-subject SD of log(CL) across occasions is
`r sprintf("%.3f", median(iov_cv$sd_log_cl))`, against the published IOV of
0.144 (14.4% CV, Table 2).

# Assumptions and deviations

* **Omega scale (% CV reading).** Table 2 labels its variability rows
  `omega^2 CL (% CV)`, `pi^2 CL (% CV)` and so on - the estimated NONMEM
  variance displayed on a percent-CV scale - but gives no footnote stating the
  conversion formula. The conventional NONMEM reading `CV% = 100 * sqrt(omega^2)`
  is used, so each variance in `ini()` is the printed percentage squared. The
  exact log-normal alternative, `CV% = 100 * sqrt(exp(omega^2) - 1)`, would give
  0.02949 / 0.03625 / 0.13966 for the three IIV terms instead of 0.02993 /
  0.03686 / 0.14977 - a difference of at most 3.6% on the omega scale, largest
  for the `V2` term where the CV is largest. No published quantity discriminates
  the two readings: the bootstrap CIs are reported on the same % CV scale, and
  the RSE column is consistent with either.
* **`sigma_add` and `sigma^2_prop` are on different scales.** Table 2's footnote a
  names the additive row as the "additive residual random error" in ug/mL and the
  proportional row as the "variance of proportional residual random error". The
  additive value is therefore used as printed (0.0210 mg/L, a standard
  deviation), while the proportional row is treated like the omegas
  (13.0% CV -> 0.130). This asymmetry is the paper's, not an assumption.
* **`sigma_add` RSE is internally inconsistent.** Table 2 prints an RSE of 0.41%
  for the additive residual error, but its bootstrap 90% CI spans
  0.000209-0.0357 around a median of 0.0224, a width implying an RSE near 100%.
  The point estimate is used regardless; the RSE is not used by the model.
* **No IIV on Q.** Table 2 prints `omega^2 Q (% CV)` as "0 FIX" and the Results
  state the IIV on intercompartmental clearance was not estimable. Rather than
  encoding a zero-variance eta - which makes the omega matrix singular and
  breaks simulation - `q` simply carries no eta. The two are mathematically
  equivalent.
* **Occasion count and definition.** Table 2 prints per-occasion shrinkage for
  Occasions 1 through 8 and the study administered 2 to 8 doses per patient, so
  eight occasion slots are encoded with one 12-hourly dose per occasion. The
  paper states the IOV variances "were assumed to be constant across occasions",
  so occasions 2-8 are `fixed()` at the occasion-1 value - the nlmixr2 spelling
  of NONMEM `$OMEGA BLOCK(1) SAME`. Setting `OCC = 0` disables IOV entirely.
* **No covariates are encoded.** Twelve covariates were screened (age, weight,
  height, sex, ECMO, CRRT, dialysis volume, ultrafiltration rate, ELWI, cardiac
  output, SOFA score, procalcitonin) plus the derived body surface area and body
  mass index. The search was *visual* - eta and kappa estimates plotted against
  covariates in Figures 3, 4 and S4-S6 - and because no trend appeared, "formal
  statistical testing was not employed". No effect size exists to encode for any
  of them. They are recorded in the model file's `covariatesDataExcluded` list so
  the provenance of the screen survives, and they carry no convention warning
  because they are never referenced in `model()`.
* **Procalcitonin units in Table 1.** Table 1 prints procalcitonin as
  "umol/liter" with a median of 8.22 and range 0.16-122. Procalcitonin is a
  13 kDa peptide reported clinically in ng/mL (= ug/L), on which 8.22 is a
  typical septic value; 8.22 umol/L would be roughly eight orders of magnitude
  above any reported concentration. The `covariatesDataExcluded` entry records
  ug/L. This affects documentation only - the covariate is not in the model.
* **Cardiac output units in Table 1.** Table 1 prints cardiac output in "liters",
  a volume, with a median of 7.49 and range 2.55-15.8. Cardiac output is a flow;
  the quoted values match the L/min scale expected in a hyperdynamic septic
  cohort. Documentation only, as above.
* **Infusion rate is set explicitly in the event table.** Doses carry
  `rate = amt / 0.5` rather than relying on a duration column, because an IV dose
  supplied without a rate is delivered as a bolus - and every AUC-based gate
  passes identically either way, so the error would be invisible. The
  closed-form check above compares against the infusion solution specifically.
* **Vss is a weak gate; the accumulation ratio is the strong one.** `Vss` = 250 L
  is just `V1 + V2` and would survive swapping the two volumes. The accumulation
  ratio of 1.57 depends on the terminal rate constant, which is a joint function
  of all four structural parameters, and is reproduced to 1.566.
* **The VPC panel shows the simulated side only.** The observed concentrations
  behind Figure 2 are not published, so the figure reproduced here is the
  prediction band alone, not the observed-versus-predicted overlay.
* **Supplemental material adds nothing to the model.** The article's supplement
  (`AAC.02273-17_zac004187006s1.pdf`, retrieved from Europe PMC) contains the
  chromatographic method, its validation, and Figures S1-S6 - time-dependent
  covariate summaries, goodness-of-fit plots, individual fits, and eta-versus-
  covariate scatter plots. It carries no NONMEM control stream and no parameter
  values beyond those in Table 2.
