---
title: "Prednisolone (Bouazza 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prednisolone (Bouazza 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Bouazza_2025_prednisolone"))$reference`
* Description: `r rxode2::rxode(readModelDb("Bouazza_2025_prednisolone"))$description`
* Article: <https://doi.org/10.1002/bcp.70103>

Bouazza and colleagues developed a population pharmacokinetic model for
prednisolone, the active metabolite of orally administered prednisone, in
patients with active systemic lupus erythematosus (SLE). The model has two
features that distinguish it from a routine one-compartment oral popPK model:

1. **The disposition is parameterised on unbound drug.** Prednisolone binds
   saturably to corticosteroid-binding globulin (CBG, high affinity / low
   capacity) and linearly to albumin (low affinity / high capacity). Total
   prednisolone therefore has markedly non-linear kinetics, while unbound
   prednisolone does not. The authors fit `CLu/F` and `Vu/F` on the unbound
   scale and recover the measured total concentration algebraically.
2. **Bioavailability falls as the dose rises.** A power function of the
   administered prednisone dose was added to `F`, which is the paper's central
   clinical message: raising the prednisone dose beyond a certain point does not
   raise effective prednisolone exposure proportionally.

# Population

The analysis pooled 66 patients with active SLE (34 juvenile-onset, 32
adult-onset) enrolled prospectively at 28 French paediatric and adult
university-hospital centres between April 2018 and January 2022
(NCT03187743). Median age was 23 years (IQR 14.2-31.2, range 6-63) and median
body weight 56.1 kg (IQR 46.8-66.5, range 17.3-113.5); 80.3% were female. The
cohort was ethnically diverse (29.2% Caucasian, 20.0% Asian, 18.5% Sub-Saharan
African, 15.4% North African, 10.8% Caribbean, remainder Turkish, Central/South
American or mixed). Renal involvement was the most common flare manifestation
(74.2%). Concomitant therapy included hydroxychloroquine (87.9%),
methylprednisolone boluses (57.6%) and mycophenolate mofetil (53.0%).

Prednisone was given orally at at least 0.5 mg/kg/day at initiation (median 0.94
mg/kg/day, IQR 0.65-1.06, range 0.38-2.3). Sampling was opportunistic during
routine care: 242 concentrations, median 3 per patient (range 1-11), drawn 5 min
to 32 h after the last dose (median 3.3 h). Thirty-five samples (14%) were below
the limit of quantification and were treated as left-censored. Baseline
characteristics are from Table 1 of the source.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Bouazza_2025_prednisolone")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Bouazza_2025_prednisolone.R`. The table
below collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `ltlag` (Tlag) | 0.17 h | Table 2, `T lag (h)`, 2% rse |
| `lka` (ka) | 1.19 1/h | Table 2, `k a (h-1)`, 1% rse |
| `lcl` (CLu/F) | 54.3 L/h per 70 kg | Table 2, `CL U /F`, 7% rse; 95% CI 48-62 (Abstract) |
| `lvc` (Vu/F) | 235 L per 70 kg | Table 2, `V U /F`, 8% rse; 95% CI 203-274 (Abstract) |
| `e_wt_cl` | 0.75 (fixed) | Table 2 covariate column `(BW/70)^0.75`; Methods 2.3 "power exponents fixed to 0.75 and 1 for CL and V" |
| `e_wt_vc` | 1 (fixed) | Table 2 covariate column `(BW/70)^1`; Methods 2.3 |
| `e_dose_fdepot` (betaFdose) | -0.28 | Table 2, `beta Fdose`, 2% rse |
| F reference dose | 80 000 nmol | Table 2 covariate column `(Dose/80000)^betaFdose`; footnote "80 000 nmol corresponds to 30 mg of prednisone" |
| `lbmax` (Bmax) | 6.77 (fixed) | Table 2, `B max`, marked `*` = fixed; Methods 2.3, set per Petersen et al. |
| `kaff` (K1) | 0.0095 L/nmol (fixed) | Table 2, `K 1 (L/nmol)`, marked `*` |
| `kns` (Kns) | 0.8 (fixed) | Table 2, `K ns`, marked `*` |
| `etalcl` | omega 0.35 -> variance 0.1225 | Table 2, `omega CL,U`, 17% rse |
| `etalbmax` | omega 0.3 (fixed) -> variance 0.09 | Table 2, `omega Bmx`, marked `*` |
| `propSd` | 0.42 | Table 2, `Residual variability, proportional`, 7% rse |
| Binding equation `Ctot = Cu * (1 + Bmax/(1 + K1*Cu) + Kns)` | n/a | Methods 2.3, displayed equation |
| Allometric covariate form `(BW/70)^exp` | n/a | Methods 2.3 and Table 2 covariate column |
| One-compartment, first-order absorption with lag | n/a | Results 3.2 first paragraph; Abstract "one-compartment open model with absorption lag time" |

## Scale conventions applied during transcription

Two conversions were required and are the most likely places for a
transcription error, so they are stated explicitly:

* **Monolix omega is a standard deviation; nlmixr2 `ini()` takes a variance.**
  Table 2's `omega CL,U = 0.35` and `omega Bmx = 0.3` are therefore encoded as
  `0.35^2 = 0.1225` and `0.3^2 = 0.09`. On the SD scale the CL random effect
  corresponds to a coefficient of variation of
  `sqrt(exp(0.1225) - 1) = 36.1%`.
* **The sign of `betaFdose`.** The published PDF's minus glyphs are lost in text
  extraction (the same file renders "-80 degrees C" as "80 C" and "h^-1" as "h1"), so
  Table 2's `beta Fdose` reads as a bare `0.28`. The exponent is negative. Three
  independent lines of evidence fix the sign, and the third is quantitative:

```{r sign-check}
# The Discussion states: for a typical 80 kg patient given 63.5 mg of
# prednisone, the model predicts an apparent unbound clearance of 75 L/h.
# Only the negative exponent reproduces that number.
mw_prednisone <- 358.43            # g/mol
dose_nmol     <- 63.5e-3 / mw_prednisone * 1e9
clu_80kg      <- 54.3 * (80 / 70)^0.75

tibble::tibble(betaFdose = c(-0.28, 0.28)) |>
  dplyr::mutate(
    F_bioav        = (dose_nmol / 80000)^betaFdose,
    apparent_CLu_F = clu_80kg / F_bioav
  ) |>
  knitr::kable(digits = 3,
               caption = "Only betaFdose = -0.28 reproduces the Discussion's 75 L/h.")
```

The other two are qualitative and agree: the Abstract states "the
bioavailability parameter was found to decrease non-linearly with the dose", and
the *What this study adds* box states "prednisolone bioavailability appears to
decrease as prednisone doses increase".

# Virtual cohort

Original observed data are not publicly available. The simulations below use
virtual populations whose covariate distributions approximate the published
trial demographics (Table 1).

```{r cohort}
set.seed(20250818)

mw_prednisone <- 358.43            # g/mol, for mg -> nmol of the dosed prodrug
mg_to_nmol    <- function(mg) mg * 1e6 / mw_prednisone

# Helper: one arm as a self-contained event table. `id_offset` keeps subject
# IDs disjoint so several arms can be bind_rows()-ed safely.
make_cohort <- function(wt, dose_mg, label, obs_times, id_offset = 0L) {
  n <- length(wt)
  subj <- tibble::tibble(
    id        = id_offset + seq_len(n),
    WT        = wt,
    treatment = label,
    dose_nmol = mg_to_nmol(dose_mg)
  )

  doses <- subj |>
    dplyr::mutate(time = 0, amt = dose_nmol, evid = 1L, cmt = "depot")

  # Observation rows point at the ODE state `central`, never at the algebraic
  # observables `Cc` / `Cu` (that would auto-inject a cmt() slot and renumber
  # the compartments).
  obs <- subj |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")

  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

obs_grid <- sort(unique(c(seq(0, 4, by = 0.05), seq(4, 24, by = 0.25))))

# Arm 1 -- the Garg & Jusko replication scenario used in the Discussion:
# a typical 80 kg patient given 63.5 mg of prednisone.
ev_gj <- make_cohort(wt = 80, dose_mg = 63.5, label = "80 kg, 63.5 mg",
                     obs_times = obs_grid, id_offset = 0L)

# Arm 2 -- typical-value dose escalation at a fixed 70 kg body weight, to
# expose the dose-dependent bioavailability.
dose_levels <- c(5, 10, 30, 60, 90, 120)
ev_dose <- dplyr::bind_rows(
  lapply(seq_along(dose_levels), function(i) {
    make_cohort(wt = 70, dose_mg = dose_levels[i],
                label = sprintf("%g mg", dose_levels[i]),
                obs_times = obs_grid, id_offset = 100L * i)
  })
)

# Arm 3 -- a 200-subject cohort matching the Table 1 weight distribution,
# dosed at the cohort median 0.94 mg/kg, for a VPC-style summary.
n_vpc  <- 200
wt_vpc <- pmin(pmax(stats::rlnorm(n_vpc, meanlog = log(56.1), sdlog = 0.2605),
                    17.3), 113.5)
ev_vpc <- make_cohort(wt = wt_vpc, dose_mg = 0.94 * wt_vpc,
                      label = "0.94 mg/kg", obs_times = obs_grid,
                      id_offset = 10000L)

# Regression guard: no duplicated (id, time, evid) triples. Note this must NOT
# be wrapped in unique() -- unique() would remove the very duplicates the
# assertion is meant to detect, making it impossible to fail.
for (ev in list(ev_gj, ev_dose, ev_vpc)) {
  stopifnot(!anyDuplicated(ev[, c("id", "time", "evid")]))
}
stopifnot(length(intersect(ev_dose$id, ev_vpc$id)) == 0)
```

The weight distribution is a log-normal truncated to the published range; its
`sdlog` is chosen so the simulated interquartile range reproduces Table 1's
46.8-66.5 kg.

```{r cohort-check}
tibble::tibble(
  statistic = c("median", "Q1", "Q3", "min", "max"),
  simulated = c(median(wt_vpc), quantile(wt_vpc, 0.25), quantile(wt_vpc, 0.75),
                min(wt_vpc), max(wt_vpc)),
  published = c(56.1, 46.8, 66.5, 17.3, 113.5)
) |>
  knitr::kable(digits = 1, caption = "Simulated vs. published body weight (Table 1).")
```

# Simulation

```{r simulate}
mod         <- readModelDb("Bouazza_2025_prednisolone")
mod_typical <- rxode2::zeroRe(mod)

# rxSolve omits the `id` column entirely when the event table holds a single
# subject, which silently breaks every downstream group-by and PKNCA formula.
# Restore it rather than assuming it is present.
solve_keep <- function(model, events) {
  out <- as.data.frame(rxode2::rxSolve(model, events = events,
                                       keep = c("WT", "treatment")))
  if (is.null(out$id)) out$id <- 1L
  out$id <- as.integer(as.character(out$id))
  out
}

sim_gj   <- solve_keep(mod_typical, ev_gj)
sim_dose <- solve_keep(mod_typical, ev_dose)
sim_vpc  <- solve_keep(mod, ev_vpc)

stopifnot(nrow(sim_gj) > 0, nrow(sim_dose) > 0, nrow(sim_vpc) > 0)
stopifnot(!anyNA(sim_gj$Cc), !anyNA(sim_dose$Cc), !anyNA(sim_vpc$Cc))
stopifnot(all(sim_vpc$Cc >= 0), all(sim_vpc$Cu >= 0))
```

Both the unbound concentration `Cu` and the observed total concentration `Cc`
are returned as columns, because rxode2 evaluates every algebraic observable at
each observation record regardless of which compartment the record points at.

# The protein-binding model

The measured quantity is total plasma prednisolone; the model's state is
unbound. The two are linked by the saturable-plus-linear binding equation of
Methods 2.3, with all three constants fixed from Petersen et al.:

$$C_{tot} = C_u \left( 1 + \frac{B_{max}}{1 + K_1 C_u} + K_{ns} \right)$$

At low concentrations CBG is far from saturated and the unbound fraction is
`1 / (1 + 6.77 + 0.8) = 11.7%`. As CBG saturates, the unbound fraction rises
towards the albumin-only limit `1 / (1 + 0.8) = 55.6%`, which is what makes
total prednisolone kinetics non-linear.

```{r binding-curve}
bmax <- 6.77; kaff <- 0.0095; kns <- 0.8

binding <- tibble::tibble(Cu = 10^seq(-1, 3.5, length.out = 400)) |>
  dplyr::mutate(
    Ctot = Cu * (1 + bmax / (1 + kaff * Cu) + kns),
    fu   = Cu / Ctot
  )

# Limiting unbound fractions, checked against the closed forms.
stopifnot(abs(min(binding$fu) - 1 / (1 + bmax + kns)) < 1e-3)
stopifnot(max(binding$fu) < 1 / (1 + kns))

ggplot(binding, aes(Ctot, 100 * fu)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 100 / (1 + bmax + kns), linetype = "dashed") +
  geom_hline(yintercept = 100 / (1 + kns), linetype = "dotted") +
  scale_x_log10() +
  labs(
    x = "Total prednisolone (nmol/L)", y = "Unbound fraction (%)",
    title = "Concentration-dependent unbound fraction",
    caption = paste(
      "Encodes the binding equation of Methods 2.3.",
      "Dashed: low-concentration limit 1/(1 + Bmax + Kns) = 11.7%.",
      "Dotted: CBG-saturated limit 1/(1 + Kns) = 55.6%."
    )
  )
```

# Dose-dependent bioavailability

This reproduces the paper's central clinical claim. Because
`F = (Dose/80000)^-0.28`, exposure rises less than proportionally with dose.

```{r figure-dose-proportionality}
auc_trapezoid <- function(time, conc) sum(diff(time) * (head(conc, -1) + tail(conc, -1)) / 2)

dose_summary <- sim_dose |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    dose_mg  = unique(as.numeric(sub(" mg", "", treatment))),
    Cmax     = max(Cc),
    AUC24_u  = auc_trapezoid(time, Cu),
    .groups  = "drop"
  ) |>
  dplyr::arrange(dose_mg) |>
  dplyr::mutate(
    F_bioav          = (mg_to_nmol(dose_mg) / 80000)^-0.28,
    AUC_per_mg       = AUC24_u / dose_mg,
    AUC_rel_to_5mg   = AUC24_u / AUC24_u[1],
    dose_rel_to_5mg  = dose_mg / dose_mg[1]
  )

dose_summary |>
  dplyr::select(treatment, F_bioav, Cmax, AUC24_u, AUC_per_mg,
                dose_rel_to_5mg, AUC_rel_to_5mg) |>
  dplyr::rename(
    "Dose"                        = treatment,
    "F"                           = F_bioav,
    "Cmax total (nmol/L)"         = Cmax,
    "AUC0-24 unbound (nmol*h/L)"  = AUC24_u,
    "AUC per mg"                  = AUC_per_mg,
    "Dose ratio vs 5 mg"          = dose_rel_to_5mg,
    "AUC ratio vs 5 mg"           = AUC_rel_to_5mg
  ) |>
  knitr::kable(digits = c(0, 3, 1, 1, 2, 1, 1),
               caption = "Dose-dependent bioavailability: a 24-fold dose increase gives a less-than-24-fold exposure increase.")

# The exposure ratio must fall below the dose ratio at every level above the
# reference, and AUC per mg must decrease monotonically.
stopifnot(all(dose_summary$AUC_rel_to_5mg[-1] < dose_summary$dose_rel_to_5mg[-1]))
stopifnot(all(diff(dose_summary$AUC_per_mg) < 0))
```

```{r figure-dose-profiles}
sim_dose |>
  dplyr::mutate(dose_lbl = factor(treatment,
                                  levels = sprintf("%g mg", sort(dose_levels)))) |>
  ggplot(aes(time, Cc, colour = dose_lbl)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Total prednisolone (nmol/L)", colour = "Prednisone dose",
       title = "Typical-value total prednisolone profiles, 70 kg",
       caption = "Dose-dependent F and saturable CBG binding both act here.")
```

# PKNCA validation

The paper reports no NCA table, so the validation targets are the quantitative
statements the authors make about the model itself. PKNCA is run on both
outputs: the unbound profile (the scale on which `CLu/F` is defined) and the
observed total profile.

```{r pknca-unbound}
nca_conc <- function(sim) {
  out <- sim |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::select(id, time, Cc, Cu, treatment)
  # Guarantee a time-zero record so PKNCA can anchor AUC from 0. Pre-dose
  # concentration is 0 for this extravascular model.
  dplyr::bind_rows(
    out,
    out |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0, Cu = 0)
  ) |>
    dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
    dplyr::arrange(id, treatment, time)
}

nca_dose <- function(ev) {
  ev |> dplyr::filter(evid == 1) |> dplyr::select(id, time, amt, treatment)
}

gj_conc <- nca_conc(sim_gj)

# Unbound: CLu/F = Dose / AUCinf on the unbound scale.
conc_u <- PKNCA::PKNCAconc(gj_conc, Cu ~ time | treatment + id,
                           concu = "nmol/L", timeu = "h")
dose_u <- PKNCA::PKNCAdose(nca_dose(ev_gj), amt ~ time | treatment + id,
                           doseu = "nmol")

intervals_u <- data.frame(start = 0, end = Inf,
                          cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE,
                          half.life = TRUE, cl.obs = TRUE)

nca_u <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_u, dose_u, intervals = intervals_u))
nca_u_tbl <- as.data.frame(nca_u$result)
knitr::kable(nca_u_tbl, digits = 3,
             caption = "PKNCA on the unbound profile, 80 kg / 63.5 mg prednisone.")
```

```{r pknca-total}
conc_t <- PKNCA::PKNCAconc(gj_conc, Cc ~ time | treatment + id,
                           concu = "nmol/L", timeu = "h")
dose_t <- PKNCA::PKNCAdose(nca_dose(ev_gj), amt ~ time | treatment + id,
                           doseu = "nmol")

nca_t <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_t, dose_t, intervals = intervals_u))
nca_t_tbl <- as.data.frame(nca_t$result)
knitr::kable(nca_t_tbl, digits = 3,
             caption = "PKNCA on the observed total profile, same subject.")
```

## Comparison against the published values

```{r nca_compare}
getp <- function(tbl, code) {
  v <- tbl$PPORRES[tbl$PPTESTCD == code]
  if (length(v) != 1L) stop("no unique PKNCA row for '", code, "'")
  v
}

clu_f_sim <- getp(nca_u_tbl, "cl.obs")
thalf_sim <- getp(nca_u_tbl, "half.life")

# Closed-form references derived from the published point estimates.
clu_80  <- 54.3 * (80 / 70)^0.75
vu_80   <- 235 * (80 / 70)
F_gj    <- (mg_to_nmol(63.5) / 80000)^-0.28

comparison <- tibble::tribble(
  ~Quantity,                                        ~Reference, ~`Reference source`,                       ~Simulated,
  "Apparent unbound clearance CLu/F (L/h)",          75.0,       "Discussion, model prediction",             clu_f_sim,
  "Apparent unbound clearance CLu/F (L/h)",          86.0,       "Discussion, Garg & Jusko observed",        clu_f_sim,
  "Unbound terminal half-life (h)",                  log(2) * vu_80 / clu_80, "log(2)*Vu/CLu from Table 2",  thalf_sim,
  "Bioavailability F at 63.5 mg",                    F_gj,       "Table 2 (Dose/80000)^-0.28",               clu_80 / clu_f_sim
) |>
  dplyr::mutate(`% diff` = 100 * (Simulated - Reference) / Reference)

comparison |>
  knitr::kable(digits = 2,
               caption = "Simulated vs. published / closed-form reference values.")
```

The model reproduces the paper's own stated prediction of **75 L/h** for a
typical 80 kg patient receiving 63.5 mg of prednisone. That number is a genuine
answer key: it appears in the Discussion, it is not one of the fitted
parameters, and it depends jointly on the clearance estimate, the allometric
exponent, the dose-to-nmol conversion and the sign and magnitude of
`betaFdose`. Reproducing it therefore checks the whole chain at once.

The 86 L/h row is the *observed* value Garg and Jusko reported in a separate
12-subject study; the authors quote it as external corroboration, and the ~13%
gap between it and the model prediction is the authors' own, not a defect
introduced here.

```{r closed-form-gates}
# Gate 1: on the unbound scale the model is linear, so AUCinf must equal
# Dose * F / CLu exactly. This simultaneously checks the dose units, the
# allometric scaling, the F term and the PKNCA setup.
auc_u_sim <- getp(nca_u_tbl, "aucinf.obs")
auc_u_ref <- mg_to_nmol(63.5) * F_gj / clu_80
stopifnot(abs(auc_u_sim - auc_u_ref) / auc_u_ref < 0.005)

# Gate 2: apparent unbound clearance must recover the Discussion's 75 L/h.
stopifnot(abs(clu_f_sim - 75) / 75 < 0.02)

# Gate 3: unbound terminal half-life must equal log(2)*Vu/CLu.
stopifnot(abs(thalf_sim - log(2) * vu_80 / clu_80) / (log(2) * vu_80 / clu_80) < 0.02)

# Gate 4: allometry. Doubling weight must scale CLu by 2^0.75 and Vu by 2.
ev_allo <- dplyr::bind_rows(
  make_cohort(35, 30, "35 kg", obs_grid, id_offset = 50000L),
  make_cohort(70, 30, "70 kg", obs_grid, id_offset = 60000L)
)
sim_allo <- solve_keep(mod_typical, ev_allo)
cl_allo <- sim_allo |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(auc = auc_trapezoid(time, Cu), WT = unique(WT), .groups = "drop") |>
  dplyr::mutate(cl_apparent = mg_to_nmol(30) / auc)
ratio_cl <- cl_allo$cl_apparent[cl_allo$WT == 70] / cl_allo$cl_apparent[cl_allo$WT == 35]
stopifnot(abs(ratio_cl - 2^0.75) / 2^0.75 < 0.02)

cat(sprintf("AUCinf unbound: sim %.1f vs closed form %.1f (%.3f%%)\n",
            auc_u_sim, auc_u_ref, 100 * (auc_u_sim - auc_u_ref) / auc_u_ref))
cat(sprintf("CLu/F: sim %.2f vs published 75 L/h (%.2f%%)\n",
            clu_f_sim, 100 * (clu_f_sim - 75) / 75))
cat(sprintf("Allometric CL ratio 70kg/35kg: %.4f vs 2^0.75 = %.4f\n",
            ratio_cl, 2^0.75))
```

# Population simulation

```{r figure-vpc}
vpc <- sim_vpc |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05), Q50 = quantile(Cc, 0.50), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot(vpc, aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.9) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Total prednisolone (nmol/L)",
       title = "Simulated total prednisolone, 200 subjects at 0.94 mg/kg prednisone",
       caption = paste("Median and 5th-95th percentiles. Comparable in shape and",
                       "spread to the prediction-corrected VPC of Figure 1."))
```

The assay was linear over 5-500 ng/mL, i.e. 13.9-1387 nmol/L for prednisolone
(molecular weight 360.44). The simulated median peak sits inside that window,
consistent with the observed data in Figure 1.

```{r assay-range}
peak_median <- max(vpc$Q50)
tibble::tibble(
  quantity = c("Simulated median Cmax (nmol/L)",
               "Assay lower limit (nmol/L)",
               "Assay upper limit (nmol/L)"),
  value    = c(peak_median, 5 / 360.44 * 1000, 500 / 360.44 * 1000)
) |>
  knitr::kable(digits = 1, caption = "Simulated peak against the published assay range.")
```

# Assumptions and deviations

- **Body-weight distribution.** Table 1 reports median, IQR and range but not a
  distributional family. A log-normal truncated to the published range is used,
  with `sdlog` chosen to reproduce the published IQR. Age, sex and ethnicity are
  not model covariates, so they are not simulated.
- **Dose units.** The model doses in **nmol of prednisone**, matching Table 2's
  `Dose` definition and its footnote ("80 000 nmol corresponds to 30 mg of
  prednisone"). Conversion in this vignette uses the prednisone molecular
  weight 358.43 g/mol; on that basis 30 mg is 83 697 nmol, so the paper's
  "80 000 nmol is about 30 mg" is a rounded statement of the reference dose, and
  80 000 nmol is used as the reference exactly as printed.
- **`betaFdose` sign.** Encoded as `-0.28`. The sign is not directly legible in
  the extracted text because the PDF's minus glyphs are dropped; it is fixed by
  the Abstract, the *What this study adds* box and the quantitative check above.
- **`F` above 1 at low doses.** Because `F` is fixed to 1 at the 80 000 nmol
  reference and modified by a negative power, doses below the reference give
  `F > 1`. This is what Table 2 specifies. `F` here is an apparent quantity
  confounded with `CLu/F` and `Vu/F`, so a value above 1 is not physically
  impossible, but simulations far below the studied dose range should be treated
  with caution: the observed range was at least 0.5 mg/kg/day.
- **Prednisone-to-prednisolone conversion.** The model does not carry a separate
  prednisone compartment. The conversion delay is absorbed into the lag time
  (as the Abstract states) and the conversion extent into `F`; the 1:1 molar
  stoichiometry means the nmol dose carries through unchanged.
- **Between-subject variability.** Only `CLu` and `Bmax` carry BSV in the final
  model. No BSV is reported on `Tlag`, `ka` or `Vu`, so none is encoded -- no
  variance was invented for them.
- **Binding constants are external.** `Bmax`, `K1` and `Kns` were fixed from
  Petersen et al. rather than estimated in this analysis (Table 2 marks all
  three with `*`). They are encoded with `fixed()` accordingly.
- **Supplementary material.** Supplemental files S1-S4 (covariate correlation
  plot, Monolix code of the final model, goodness-of-fit plots, BLQ predictive
  check) are referenced by the paper but were not available on disk. Every value
  used here comes from Table 2 and the Methods text of the main article, which
  are complete for the final model; the Monolix code in S2 would provide an
  independent confirmation of the equation forms but is not required to
  reproduce them.
- **Validation targets.** The paper reports no NCA table, so the comparison
  table uses the authors' own quantitative statement (75 L/h) plus closed-form
  identities derived from the Table 2 point estimates. No parameter was tuned.
