---
title: "Iadademstat (ORY-1001) in vitro to in vivo QSP model (Bouhaddou 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Iadademstat (ORY-1001) in vitro to in vivo QSP model (Bouhaddou 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Bouhaddou 2020 builds a semimechanistic PK/PD model of iadademstat
(ORY-1001), a covalent inhibitor of the histone demethylase LSD1 (KDM1A), in
the NCI-H510A small-cell lung cancer (SCLC) cell line. The model is trained
almost entirely on cell-culture data and then used to *predict* xenograft
tumour growth in mice. The authors built two models, and the paper contributes
two files to `nlmixr2lib`:

```{r models}
mod_vitro <- rxode2::rxode(readModelDb("Bouhaddou_2020_iadademstat_invitro"))
mod_vivo <- rxode2::rxode(readModelDb("Bouhaddou_2020_iadademstat_mouse"))
```

```{r model-table, echo = FALSE}
tibble::tibble(
  file = c(
    "Bouhaddou_2020_iadademstat_invitro",
    "Bouhaddou_2020_iadademstat_mouse"
  ),
  role = c(
    "In vitro PD model: constant drug concentration -> LSD1 target engagement -> GRP mRNA -> proliferating / quiescent cell switch (Figures 2-3)",
    "In vivo PK/PD model: oral two-compartment PK with dose-dependent bioavailability, unbound plasma concentration driving the same PD model; only kP re-estimated (Figures 4-5)"
  )
) |>
  dplyr::rename("nlmixr2lib model" = file, "Role" = role) |>
  knitr::kable(caption = "The two model files contributed by Bouhaddou 2020.")
```

* Citation: `r mod_vivo$reference`
* Article: <https://doi.org/10.1111/cts.12727>
* PubMed Central: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7070804/>

The paper's display equations (Eqs. 1-20) give the structure; the parameter
values are in Supplementary Tables S1 (initial conditions) and S2
(parameters). The authors also deposited their MATLAB code, the fitted
parameter files and the experimental data as a supplementary zip
(`code_Bouhaddouetal_pub`). The deposit settles one printed-equation defect
(Eq. 1, see *Assumptions and deviations*) and supplies the observed data that
the figure replications below are compared against.

# Population

* **In vitro:** NCI-H510A cells. Target engagement was measured by an
  ELISA-based free/total LSD1 chemoprobe assay after 24 h exposure to 0.6, 3 or
  15 nM followed by 3 days of wash-out; GRP mRNA by TaqMan RT-qPCR after
  continuous 0.1, 1 or 10 nM for up to 7 days and after 5 nM pulsed
  (3 days on / 7 off) or continuous (10 days) exposure; growth and viability by
  alamarBlue in 96-well plates seeded at 8000 cells per well, over 10 days.
  Three biological replicates per condition.
* **In vivo:** female athymic nude mice, 7-8 weeks old, implanted
  subcutaneously with 5 million NCI-H510A cells. PK was measured in tumour-free
  mice after single oral doses (3 mice per time point); efficacy in five arms
  of 13-15 mice (vehicle, 20 and 40 ug/kg once daily 5-on/2-off, 200 ug/kg
  weekly, 400 ug/kg on days 7, 16 and 23). Mouse plasma fraction unbound was
  0.75.

The same information is available programmatically via the `population`
metadata, e.g. `readModelDb("Bouhaddou_2020_iadademstat_mouse")()$population`.

# Source trace

Every `ini()` value carries an in-file comment naming its source. The table
collects them. "Code" refers to the deposited MATLAB files.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Depot, central, peripheral ODEs | n/a | Eqs. 1-3 (p. 420-421); `RunModelVivo.m` |
| Dose-dependent bioavailability `F = D / (D50 + D)` | n/a | Eq. 1 as implemented in `RunModelVivo.m` (`Xup = (X0/(D50+X0))*X0`) |
| `Cc = central / V` | n/a | Eq. 5 |
| Unbound active concentration `roc = Cc * fu * 1000 / MW` (nM) | n/a | Eq. 4; `RunModelVivo.m` (`fac_ng2nM`) |
| `d/dt(roc) = 0` (in vitro) | n/a | Eq. 6 |
| Bound LSD1 ODE, unbound LSD1, TE | n/a | Eqs. 7-9 (p. 422) |
| GRP mRNA ODE, `kmax = -m*GRP + b`, `m = b - kdeg` | n/a | Eqs. 10-12 (p. 422) |
| Proliferating / quiescent ODEs and switching rates | n/a | Eqs. 13-20 (p. 422) |
| `lka` | log(0.88205) 1/h | Table S2 |
| `lcl` | log(117.66) mL/h | Table S2 ('kCL'); code `CL = CLml/V` |
| `lvc` | log(577.49) mL | Table S2 |
| `lk12`, `lk21` | log(0.47004), log(0.25469) 1/h | Table S2 |
| `ld50` | log(1123.9) ng | Table S2 |
| `fu` | 0.75 (measured) | Table S2; Results p. 424 |
| `mw` | 230.4 g/mol | Code `rateconstants_vivo_BEST.txt` |
| `ki` | 19 nM (measured) | Table S2 |
| `lkinact` | log(0.87052) 1/h | Table S2 |
| `lvmax_lsd1b`, `lkm_lsd1b` | log(0.03515) nM/h, log(0.9028) nM | Table S2 |
| `llsd1tot` | log(3.0833) nM | Table S2 |
| `lte50`, `hill_te` | log(26.513) %, 2.0819 | Table S2 |
| `lkdeg_grp`, `b_grp` | log(0.025567) 1/h, 0.083048 1/h | Table S2 |
| `lkp` | log(0.023) in vitro; log(0.0037) in vivo (1/h) | Table S2 |
| `lk50p` | log(100000) | Table S2 |
| `lkmaxpq`, `lkmaxqp` | log(0.5932), log(4.4908) 1/h | Table S2 |
| `k50pq`, `k50qp` | 0.8836, 0.99897 | Table S2 |
| `hill_pq`, `hill_qp` | 3.8313, 37.455 | Table S2 |
| `bs` | 0.033328 | Table S2 |
| `prolif0` | 8000 cells in vitro; 70 mm^3 in vivo | Table S1 |
| `grp(0)` | 1 | Table S1 |

# Reproduction gate: an independent implementation of the authors' code

The deposited `RunModelVivo.m` is simulated here a second time with `deSolve`,
following the MATLAB driver literally: 35 consecutive 24-hour integrations,
the depot *overwritten* with the dose at the start of each dosing day, and the
bioavailability correction `X0 <- X0 * X0 / (D50 + X0)` applied to the depot at
the start of every 24-hour chunk (`RunDosingRegVivo.m`). The `nlmixr2lib`
model instead adds each dose through a single `f(depot)`. The two differ only
by the negligible depot residue left after 24 h (`exp(-0.882 * 24)` ~ 6e-10),
so they must agree to solver precision.

```{r matlab-reimplementation}
# Parameter vector in the order of rateconstants_vivo_BEST.txt.
k_vivo <- c(
  Ki = 19, kinact = 0.87052, vm = 0.03515, km = 0.9028, LSD1_0 = 3.0833,
  k50 = 26.513, n = 2.0819, kd = 0.025567, b = 0.083048, kP = 0.0037,
  k50P = 1e5, kmaxPQ = 0.5932, kmaxQP = 4.4908, k50PQ = 0.8836,
  k50QP = 0.99897, nPQ = 3.8313, nQP = 37.455, bs = 0.033328,
  ka = 0.88205, k12 = 0.47004, k21 = 0.25469, V = 577.49, D50 = 1123.9,
  CLml = 117.66, fu = 0.75, MW = 230.4
)

# Right-hand side of RunModelVivo.m's createODEs(); the ROc ODE is the
# authors' own (fac_ng2nM times dQc/dt).
matlab_rhs <- function(t, y, k) {
  ROc <- y[1]; LSD1B <- y[2]; GRP <- y[3]; P <- y[4]; Q <- y[5]
  X <- y[6]; Qc <- y[7]; Qp <- y[8]
  CL <- k[["CLml"]] / k[["V"]]
  dX <- -k[["ka"]] * X
  dQc <- k[["ka"]] * X - CL * Qc - k[["k12"]] * Qc + k[["k21"]] * Qp
  dQp <- k[["k12"]] * Qc - k[["k21"]] * Qp
  fac <- (1000 / k[["MW"]]) / k[["V"]] * k[["fu"]]
  dROc <- dQc * fac
  LSD1U <- k[["LSD1_0"]] - LSD1B
  dLSD1B <- -LSD1B * (k[["vm"]] / (k[["km"]] + LSD1B)) +
    k[["kinact"]] * ROc / (k[["Ki"]] + ROc) * LSD1U
  TE <- LSD1B / k[["LSD1_0"]] * 100
  m <- k[["b"]] - k[["kd"]]
  kmax <- -m * GRP + k[["b"]]
  dGRP <- kmax * (k[["k50"]]^k[["n"]] / (k[["k50"]]^k[["n"]] + TE^k[["n"]])) -
    k[["kd"]] * GRP
  vP <- k[["kP"]] * (1 - P / (k[["k50P"]] + P)) * P
  BM <- abs(1 - GRP)
  vPQ <- k[["kmaxPQ"]] * BM^k[["nPQ"]] / (k[["k50PQ"]]^k[["nPQ"]] + BM^k[["nPQ"]]) * P
  BMQP <- 1 - BM
  vQP <- k[["kmaxQP"]] *
    (BMQP^k[["nQP"]] / (k[["k50QP"]]^k[["nQP"]] + BMQP^k[["nQP"]]) + k[["bs"]]) * Q
  list(c(dROc, dLSD1B, dGRP, vP - vPQ + vQP, vPQ - vQP, dX, dQc, dQp))
}

# RunDosingRegVivo.m: day-by-day driver; dose_ugkg converted with a 0.025 kg mouse.
matlab_regimen <- function(dose_ugkg, dose_days, k, total_days = 35) {
  y <- c(0, 0, 1, 70, 0, 0, 0, 0)
  out <- list()
  for (i in seq_len(total_days)) {
    if (i %in% dose_days) y[6] <- dose_ugkg * 1e3 * 0.025
    y[6] <- y[6] / (k[["D50"]] + y[6]) * y[6]
    s <- deSolve::lsoda(y, seq(0, 24, by = 1), matlab_rhs, k, rtol = 1e-10, atol = 1e-12)
    out[[i]] <- data.frame(
      time = s[-nrow(s), 1] + (i - 1) * 24,
      Cc = s[-nrow(s), 8] / k[["V"]],
      TE = s[-nrow(s), 3] / k[["LSD1_0"]] * 100,
      grp = s[-nrow(s), 4],
      tumor_vol = s[-nrow(s), 5] + s[-nrow(s), 6]
    )
    y <- s[nrow(s), -1]
  }
  dplyr::bind_rows(out)
}

# MATLAB day indices of dosing (RunDosingRegVivo.m, `dayson`).
regimens <- tibble::tibble(
  arm = c(
    "Vehicle", "20 ug/kg QD 5on/2off", "40 ug/kg QD 5on/2off",
    "200 ug/kg weekly", "400 ug/kg days 7, 16, 23"
  ),
  dose_ugkg = c(0, 20, 40, 200, 400),
  dose_days = list(
    integer(), c(14:18, 21:25, 28:32, 35), c(14:18, 21:25, 28:32, 35),
    c(7, 14, 21, 28), c(7, 16, 23)
  )
) |>
  dplyr::mutate(arm = factor(arm, levels = arm))

# nlmixr2lib event table: dose on MATLAB day index i is given at (i - 1) * 24 h.
# Observations are on ODE states (prolif), so every algebraic output is returned.
regimen_events <- function(dose_ugkg, dose_days, arm_id) {
  obs <- data.frame(id = arm_id, time = seq(0, 35 * 24, by = 1), evid = 0, amt = 0, cmt = "prolif")
  if (length(dose_days) == 0L || dose_ugkg == 0) {
    return(obs)
  }
  dose <- data.frame(
    id = arm_id, time = (dose_days - 1) * 24, evid = 1,
    amt = dose_ugkg * 1e3 * 0.025, cmt = "depot"
  )
  dplyr::bind_rows(dose, obs) |> dplyr::arrange(time, dplyr::desc(evid))
}

ev_vivo <- dplyr::bind_rows(lapply(seq_len(nrow(regimens)), function(i) {
  regimen_events(regimens$dose_ugkg[i], regimens$dose_days[[i]], i)
}))
sim_vivo <- rxode2::rxSolve(mod_vivo, ev_vivo, rtol = 1e-10, atol = 1e-12) |>
  as.data.frame() |>
  dplyr::mutate(arm = regimens$arm[id])

ref_vivo <- dplyr::bind_rows(lapply(seq_len(nrow(regimens)), function(i) {
  matlab_regimen(regimens$dose_ugkg[i], regimens$dose_days[[i]], k_vivo) |>
    dplyr::mutate(arm = regimens$arm[i])
}))

cmp_vivo <- dplyr::inner_join(sim_vivo, ref_vivo, by = c("arm", "time"), suffix = c("", "_ref"))
stopifnot(nrow(cmp_vivo) == 5 * 35 * 24)
repro <- cmp_vivo |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    tumor_vol = max(abs(tumor_vol / tumor_vol_ref - 1)),
    grp = max(abs(grp - grp_ref)),
    TE = max(abs(TE - TE_ref)),
    Cc = max(abs(Cc - Cc_ref)) / max(max(Cc_ref), 1e-12),
    .groups = "drop"
  )
knitr::kable(
  repro |>
    dplyr::rename(
      "Arm" = arm, "tumor_vol (max rel. diff)" = tumor_vol,
      "GRP (max abs. diff)" = grp, "TE (max abs. diff, %)" = TE,
      "Cc (max diff / Cmax)" = Cc
    ),
  digits = 8,
  caption = "nlmixr2lib model vs an independent deSolve run of the authors' MATLAB driver."
)
# Both sides solve the same deterministic ODE with the same parameters, so the
# difference is numerical error only and a tight bound is correct (a
# mis-transcribed parameter or a wrong dose conversion moves these by >1%).
stopifnot(
  max(repro$tumor_vol) < 1e-4,
  max(repro$grp) < 1e-4,
  max(repro$TE) < 1e-3,
  max(repro$Cc) < 1e-4
)
```

# In vitro model: Figures 2 and 3

In the cell-culture experiments the drug concentration is constant while it is
on (Eq. 6). Exposure starts with a dose of the concentration (nM) into `roc`
and ends with a replacement event (`evid = 5`, `amt = 0`), which is how
`RunModelVitro.m` models the wash-off (`y2(1) = 0`).

```{r vitro-helpers}
# One well: exposure to conc_nM from time 0 to t_off (h), observed on `obs_times`.
vitro_events <- function(conc_nM, t_off, obs_times, well) {
  ev <- data.frame(id = well, time = obs_times, evid = 0, amt = 0, cmt = "prolif")
  if (conc_nM > 0) {
    ev <- dplyr::bind_rows(
      data.frame(id = well, time = 0, evid = 1, amt = conc_nM, cmt = "roc"),
      ev
    )
    if (is.finite(t_off)) {
      ev <- dplyr::bind_rows(
        ev,
        data.frame(id = well, time = t_off, evid = 5, amt = 0, cmt = "roc")
      )
    }
  }
  dplyr::arrange(ev, time, dplyr::desc(evid == 0))
}
```

## Figure 2a -- target engagement after a 24-hour pulse

```{r fig2a}
te_obs <- tibble::tribble(
  ~conc, ~time, ~TE,
  0.6, 24, 58.24, 0.6, 48, 11.81, 0.6, 72, 12.97, 0.6, 96, 10.20,
  3, 24, 81.07, 3, 48, 65.96, 3, 72, 47.79, 3, 96, 34.19,
  15, 24, 84.76, 15, 48, 68.99, 15, 72, 65.23, 15, 96, 50.91
)
concs_te <- c(0.6, 3, 15)
ev_te <- dplyr::bind_rows(lapply(seq_along(concs_te), function(i) {
  vitro_events(concs_te[i], 24, seq(0, 96, by = 1), i)
}))
sim_te <- rxode2::rxSolve(mod_vitro, ev_te) |>
  as.data.frame() |>
  dplyr::mutate(conc = concs_te[id])

ggplot(sim_te, aes(time / 24, TE, colour = factor(conc))) +
  geom_line() +
  geom_point(data = te_obs, size = 2) +
  labs(
    x = "Time (days)", y = "Target engagement (%)", colour = "ORY-1001 (nM)",
    caption = "Replicates Figure 2a of Bouhaddou 2020 (24 h exposure, then wash-off). Points: observed means."
  )
```

## Figure 2b and 2c -- GRP mRNA under continuous and pulsed exposure

```{r fig2bc}
grp_obs <- tibble::tribble(
  ~conc, ~time, ~grp,
  0.1, 24, 1.167, 0.1, 96, 0.744, 0.1, 168, 0.936,
  1, 24, 0.813, 1, 96, 0.274, 1, 168, 0.250,
  10, 24, 0.782, 10, 96, 0.227, 10, 168, 0.219
)
concs_grp <- c(0.1, 1, 10)
ev_grp <- dplyr::bind_rows(lapply(seq_along(concs_grp), function(i) {
  vitro_events(concs_grp[i], Inf, seq(0, 168, by = 2), i)
}))
sim_grp <- rxode2::rxSolve(mod_vitro, ev_grp) |>
  as.data.frame() |>
  dplyr::mutate(conc = concs_grp[id])

ggplot(sim_grp, aes(time / 24, grp, colour = factor(conc))) +
  geom_line() +
  geom_point(data = grp_obs, size = 2) +
  labs(
    x = "Time (days)", y = "GRP mRNA (relative to vehicle)", colour = "ORY-1001 (nM)",
    caption = "Replicates Figure 2b of Bouhaddou 2020 (continuous exposure). Points: observed means."
  )

# Figure 2c: 5 nM for 3 days then 7 days off, versus 10 days continuous.
ev_wo <- dplyr::bind_rows(
  vitro_events(5, 72, seq(0, 240, by = 2), 1),
  vitro_events(5, Inf, seq(0, 240, by = 2), 2)
)
sim_wo <- rxode2::rxSolve(mod_vitro, ev_wo) |>
  as.data.frame() |>
  dplyr::mutate(schedule = c("3 days on, 7 off", "10 days on")[id])
wo_day10 <- sim_wo |>
  dplyr::filter(time == 240) |>
  dplyr::select(schedule, grp) |>
  dplyr::mutate(observed = c(1.10, 0.19)[match(schedule, c("3 days on, 7 off", "10 days on"))])
knitr::kable(
  wo_day10 |> dplyr::rename("Schedule" = schedule, "Simulated GRP, day 10" = grp, "Observed GRP, day 10" = observed),
  digits = 3,
  caption = "Figure 2c bar plot: GRP mRNA at day 10 after 5 nM pulsed or continuous exposure."
)
# Structural: a 3-day pulse recovers GRP to baseline by day 10, continuous
# exposure keeps it suppressed (Results p. 423).
stopifnot(
  wo_day10$grp[wo_day10$schedule == "3 days on, 7 off"] > 0.9,
  wo_day10$grp[wo_day10$schedule == "10 days on"] < 0.3
)
```

## Figure 3 -- cell growth and viability

```{r fig3}
# Figure 3a: drug-free growth from 8000 seeded cells.
growth_obs <- tibble::tibble(
  time = c(0, 24, 48, 96, 168, 240),
  cells = c(8501, 24708, 32240, 48168, 130760, 231063)
)
sim_growth <- rxode2::rxSolve(mod_vitro, vitro_events(0, Inf, growth_obs$time, 1)) |>
  as.data.frame()
growth_cmp <- dplyr::mutate(growth_obs, simulated = sim_growth$cell_count)
knitr::kable(
  growth_cmp |> dplyr::rename("Time (h)" = time, "Observed cells" = cells, "Simulated cells" = simulated),
  digits = 0, caption = "Figure 3a: drug-free NCI-H510A growth (observed means from the deposited data)."
)

# Figure 3b: day-10 viability versus concentration, relative to no drug.
cv_obs <- tibble::tibble(
  conc = c(1000, 333.33, 111.11, 37.04, 12.35, 4.115, 1.372, 0.4572, 0.1524),
  viability = 100 * c(109460, 97219, 96041, 90953, 91565, 92858, 137010, 174883, 211239) / 235795
)
conc_grid <- 10^seq(-2, 4, length.out = 41)
ev_cv <- dplyr::bind_rows(lapply(seq_along(conc_grid), function(i) {
  vitro_events(conc_grid[i], Inf, 240, i)
}))
ctrl_10d <- growth_cmp$simulated[growth_cmp$time == 240]
sim_cv <- rxode2::rxSolve(mod_vitro, ev_cv) |>
  as.data.frame() |>
  dplyr::mutate(conc = conc_grid[id], viability = 100 * cell_count / ctrl_10d)

ggplot(sim_cv, aes(conc, viability)) +
  geom_line() +
  geom_point(data = cv_obs, shape = 8) +
  scale_x_log10() +
  labs(
    x = "ORY-1001 (nM)", y = "Viability at day 10 (% of no drug)",
    caption = "Replicates Figure 3b of Bouhaddou 2020. Asterisks: observed means."
  )

# Figure 3c: 5 nM for 3, 5 or 7 days, then off until day 10.
don <- c(3, 5, 7)
ev_pulse <- dplyr::bind_rows(lapply(seq_along(don), function(i) {
  vitro_events(5, 24 * don[i], 240, i)
}))
sim_pulse <- rxode2::rxSolve(mod_vitro, ev_pulse) |>
  as.data.frame() |>
  dplyr::transmute(
    schedule = paste(don[id], "days on"),
    simulated = 100 * cell_count / ctrl_10d,
    # Blue 'Sim' bars of Figure 3c, digitised by the maintainers (+/- 2%).
    published_sim = c(66, 52, 43)[id],
    observed = c(48.9, 35.3, 26.5)[id]
  )
knitr::kable(
  sim_pulse |>
    dplyr::rename(
      "Schedule (5 nM)" = schedule, "Simulated viability (%)" = simulated,
      "Published simulation, Fig. 3c (%)" = published_sim,
      "Observed viability (%)" = observed
    ),
  digits = 1, caption = "Figure 3c: day-10 viability after pulsed 5 nM exposure."
)
```

The model over-predicts pulsed-exposure viability by about 16 percentage
points relative to the observations, and the published simulation bars of
Figure 3c show the same gap, so this is a property of the authors' fit, not of
the transcription.

```{r fig3-gates}
stopifnot(
  # Pulsed viability reproduces the authors' own simulated bars.
  all(abs(sim_pulse$simulated - sim_pulse$published_sim) < 4),
  # Drug-free growth: within 20% of the observed day-10 cell count.
  abs(ctrl_10d / 231063 - 1) < 0.2,
  # Cytostatic plateau: viability saturates near 40-50% at high concentration.
  abs(sim_cv$viability[which.max(sim_cv$conc)] - 46) < 15
)
```

# In vivo model: Figures 4 and 5

## Figure 4 -- single-dose oral PK in mice

The deposited PK sheet labels its dose column 1, 2 and 500; `Plot_PK.m`
simulates those samples as 20, 40 and 10000 ug/kg, i.e. the column is a
multiple of 20 ug/kg. The dose is entered in ng (`dose_ugkg * 1000 * 0.025`).

```{r fig4}
pk_obs <- tibble::tribble(
  ~dose_ugkg, ~time, ~Cc,
  20, 1, 0.164, 20, 1, 0.136, 20, 1, 0.0988, 20, 6, 0.043, 20, 6, 0.0217,
  20, 24, 0.0135, 20, 24, 0.0102,
  40, 1, 0.347, 40, 1, 0.392, 40, 1, 0.382, 40, 6, 0.158, 40, 6, 0.201,
  40, 6, 0.229, 40, 24, 0.0532, 40, 24, 0.136, 40, 24, 0.0437,
  10000, 1, 176.87, 10000, 4, 155.24, 10000, 1, 210.28, 10000, 4, 144.85,
  10000, 1, 216.74, 10000, 4, 167.96, 10000, 0.25, 68.43, 10000, 2, 118.33,
  10000, 8, 89.33, 10000, 0.25, 82.89, 10000, 2, 199.27, 10000, 8, 93.54,
  10000, 0.25, 63.72, 10000, 2, 169.45, 10000, 8, 71.69, 10000, 0.5, 130.6,
  10000, 4, 86.66, 10000, 24, 19.14, 10000, 0.5, 136.21, 10000, 4, 104.15,
  10000, 24, 21.46, 10000, 0.5, 78.92, 10000, 4, 95.26, 10000, 24, 34.38
)
pk_doses <- c(20, 40, 10000)
pk_times <- sort(unique(c(seq(0, 2, by = 0.05), seq(2.25, 24, by = 0.25), seq(25, 168, by = 1))))
ev_pk <- dplyr::bind_rows(lapply(seq_along(pk_doses), function(i) {
  dplyr::bind_rows(
    data.frame(id = i, time = 0, evid = 1, amt = pk_doses[i] * 1e3 * 0.025, cmt = "depot"),
    data.frame(id = i, time = pk_times, evid = 0, amt = 0, cmt = "central")
  )
}))
sim_pk <- rxode2::rxSolve(mod_vivo, ev_pk) |>
  as.data.frame() |>
  dplyr::mutate(dose_ugkg = pk_doses[id], dose_ng = dose_ugkg * 1e3 * 0.025)

ggplot(dplyr::filter(sim_pk, time <= 24), aes(time, Cc / dose_ugkg, colour = factor(dose_ugkg))) +
  geom_line() +
  geom_point(data = pk_obs) +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Dose-normalised plasma concentration (ng/mL per ug/kg)",
    colour = "Dose (ug/kg)",
    caption = "Replicates Figure 4b of Bouhaddou 2020: the curves are parallel but do not superimpose (dose-dependent bioavailability)."
  )

pk_pred_obs <- pk_obs |>
  dplyr::left_join(
    dplyr::select(sim_pk, dose_ugkg, time, pred = Cc),
    by = c("dose_ugkg", "time")
  )
stopifnot(!anyNA(pk_pred_obs$pred))
pk_fit <- pk_pred_obs |>
  dplyr::group_by(dose_ugkg) |>
  dplyr::summarise(gmr = exp(mean(log(pred / Cc))), .groups = "drop")
knitr::kable(
  pk_fit |> dplyr::rename("Dose (ug/kg)" = dose_ugkg, "Geometric mean predicted / observed" = gmr),
  digits = 2, caption = "Figure 4c: model predictions against the deposited plasma concentrations."
)
stopifnot(all(abs(log(pk_fit$gmr)) < log(1.5)))
```

## PKNCA: the dose-dependent bioavailability

Bouhaddou 2020 reports no NCA table, so there is no published NCA to compare
against. PKNCA is instead used to check the one PK feature that is easy to get
wrong in translation, the saturable bioavailability. `Dose / AUCinf` is the
apparent clearance `CL/F`; multiplied by the model's `F = D / (D50 + D)` it
must return `CL = 117.66 mL/h` at every dose, while the apparent clearance
itself falls steeply with dose.

```{r pknca}
conc_pk <- sim_pk |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(treatment = paste(dose_ugkg, "ug/kg")) |>
  dplyr::select(id, time, Cc, treatment)
dose_pk <- ev_pk |>
  dplyr::filter(evid == 1) |>
  dplyr::mutate(treatment = paste(pk_doses[id], "ug/kg")) |>
  dplyr::select(id, time, amt, treatment)

o_conc <- PKNCA::PKNCAconc(conc_pk, Cc ~ time | treatment + id)
o_dose <- PKNCA::PKNCAdose(dose_pk, amt ~ time | treatment + id)
o_data <- PKNCA::PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(
    start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
    aucinf.obs = TRUE, cl.obs = TRUE, half.life = TRUE
  )
)
nca_res <- PKNCA::pk.nca(o_data)

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::mutate(
    dose_ng = as.numeric(sub(" ug/kg", "", treatment)) * 1e3 * 0.025,
    F = dose_ng / (1123.9 + dose_ng),
    cl_recovered = cl.obs * F
  ) |>
  dplyr::arrange(dose_ng)
knitr::kable(
  nca_wide |>
    dplyr::select(treatment, cmax, tmax, aucinf.obs, half.life, cl.obs, F, cl_recovered) |>
    dplyr::rename(
      "Dose" = treatment, "Cmax (ng/mL)" = cmax, "Tmax (h)" = tmax,
      "AUC0-inf (ng*h/mL)" = aucinf.obs, "t1/2 (h)" = half.life,
      "CL/F (mL/h)" = cl.obs, "F" = F, "CL/F x F (mL/h)" = cl_recovered
    ),
  digits = 3, caption = "PKNCA on the simulated single oral doses (0-168 h)."
)
stopifnot(all(abs(nca_wide$cl_recovered / 117.66 - 1) < 0.02))
```

## Figure 5 -- predicted tumour growth under five regimens

```{r fig5}
tg_obs <- dplyr::bind_rows(
  tibble::tibble(
    arm = "Vehicle", day = c(7, 10, 14, 17, 21, 24, 28, 31, 35),
    vol = c(115.0, 122.9, 153.7, 224.7, 390.5, 644.0, 910.9, 1149.6, 1618.1)
  ),
  tibble::tibble(
    arm = "20 ug/kg QD 5on/2off", day = c(7, 10, 14, 17, 21, 24, 28, 31, 35),
    vol = c(75.7, 96.9, 153.1, 244.6, 438.2, 782.9, 1145.9, 1339.3, 1820.2)
  ),
  tibble::tibble(
    arm = "40 ug/kg QD 5on/2off", day = c(7, 10, 14, 17, 21, 24, 28, 31, 35),
    vol = c(76.9, 97.7, 154.5, 226.7, 323.5, 473.4, 682.4, 740.6, 1065.1)
  ),
  tibble::tibble(
    arm = "200 ug/kg weekly", day = c(6, 9, 13, 16, 21, 23, 27, 30),
    vol = c(79.4, 97.5, 146.6, 203.4, 459.1, 548.4, 864.6, 1162.6)
  ),
  tibble::tibble(
    arm = "400 ug/kg days 7, 16, 23", day = c(6, 9, 13, 16, 21, 23, 27, 30),
    vol = c(81.9, 98.2, 129.2, 147.0, 309.1, 424.0, 634.3, 807.9)
  )
) |>
  dplyr::mutate(arm = factor(arm, levels = levels(regimens$arm)))

ggplot(sim_vivo, aes(time / 24, tumor_vol)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(data = tg_obs, aes(day, vol), shape = 8) +
  facet_wrap(~arm, nrow = 1) +
  labs(
    x = "Time (days)", y = "Tumour volume (mm^3)",
    caption = "Replicates Figure 5b of Bouhaddou 2020. Line: model prediction; asterisks: observed means."
  )

sim_vivo |>
  dplyr::select(time, arm, Cc, roc, TE, grp) |>
  tidyr::pivot_longer(c(Cc, roc, TE, grp)) |>
  dplyr::mutate(name = factor(name, levels = c("Cc", "roc", "TE", "grp"))) |>
  ggplot(aes(time / 24, value, colour = arm)) +
  geom_line() +
  facet_wrap(~name, ncol = 1, scales = "free_y") +
  labs(
    x = "Time (days)", y = NULL, colour = NULL,
    caption = "Replicates Figure 5a: plasma (ng/mL), unbound tumour (nM), TE (%) and GRP mRNA."
  ) +
  theme(legend.position = "bottom")

# Figure 5c: simulated versus observed at the observation times.
tg_cmp <- tg_obs |>
  dplyr::left_join(
    sim_vivo |> dplyr::mutate(day = time / 24) |> dplyr::select(arm, day, tumor_vol),
    by = c("arm", "day")
  )
stopifnot(!anyNA(tg_cmp$tumor_vol))

# Figure 5d: ratio of the areas under the tumour growth curves, computed with
# the authors' convention (Plot_TG_Predictions.m): the simulated area runs from
# day 0 to day 35 (day 30 for the two weekly arms), the observed area over the
# observed time points only.
trap <- function(x, y) sum(diff(x) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)
sim_end <- c(35, 35, 35, 30, 30)
augic <- tg_cmp |>
  dplyr::group_by(arm) |>
  dplyr::summarise(auc_exp = trap(day, vol), .groups = "drop") |>
  dplyr::mutate(
    auc_sim = vapply(seq_along(arm), function(i) {
      s <- sim_vivo[sim_vivo$arm == arm[i] & sim_vivo$time <= sim_end[i] * 24, ]
      trap(s$time / 24, s$tumor_vol)
    }, numeric(1)),
    ratio = auc_sim / auc_exp,
    # Figure 5d bar heights, digitised by the maintainers (+/- 0.03).
    published = c(1.05, 0.91, 1.04, 0.78, 1.01)
  )
knitr::kable(
  augic |>
    dplyr::select(arm, ratio, published) |>
    dplyr::rename("Arm" = arm, "AUGIC ratio (this model)" = ratio, "AUGIC ratio, Fig. 5d" = published),
  digits = 2, caption = "Replicates Figure 5d: simulated / observed area under the tumour growth curve."
)
stopifnot(
  # Reproduces the published bars; a wrong kP, dose conversion or PK
  # parameter moves these ratios by 0.1 or more.
  all(abs(augic$ratio - augic$published) < 0.06),
  # The drug effect is real: 40 ug/kg 5on/2off ends well below vehicle.
  tg_cmp$tumor_vol[tg_cmp$arm == "40 ug/kg QD 5on/2off" & tg_cmp$day == 35] <
    0.7 * tg_cmp$tumor_vol[tg_cmp$arm == "Vehicle" & tg_cmp$day == 35]
)
```

# Assumptions and deviations

- **Eq. 1 as printed is defective.** It reads `dX/dt = -ka * D/(D50 + D) * D`,
  which does not balance against Eq. 2 (`+ka * X`). The deposited
  `RunModelVivo.m` implements a first-order depot, `dX = -ka*X`, and scales the
  administered amount once at dosing by `D / (D50 + D)`
  (`Xup = (X0/(D50+X0))*X0`). The model follows the code: a dose-dependent
  bioavailability `f(depot) = podo(depot) / (D50 + podo(depot))`, with the
  dose in ng. The reproduction gate above confirms the two are equivalent.
- **`kCL` is a clearance, not a rate constant.** Table S2 lists
  `kCL = 117.66 1/h`, but the code divides it by `V` (`CL = CLml/V`), and the
  parameter file names it `CLml`. It is encoded as `CL = 117.66 mL/h`
  (elimination rate constant 0.204 1/h).
- **Dose units.** `D50` is "a.u." in Table S2; the code applies it to the dose
  in ng after converting ug/kg with a fixed 0.025 kg mouse. Doses must
  therefore be entered in ng (`ug/kg * 25`).
- **Unbound concentration in nM.** Eq. 4 (`ROc = CPL * fu`) is completed with
  the ng/mL to nM conversion the code uses, `1000 / MW` with
  `MW = 230.4 g/mol` (from the deposited parameter file; not printed in the
  paper).
- **k50P in vivo.** The Results say kP and k50P were re-estimated for the
  xenograft, but Table S2 lists k50P as "Same" and the deposited in vivo
  parameter file keeps 100000; the model uses 100000 (in mm^3 it is so large
  that growth is effectively exponential over 35 days).
- **Point estimates vs intervals.** Several Table S2 point estimates (e.g.
  vmax 0.03515 with 95% CI 0.0576-0.1914) lie outside the tabulated interval.
  The point estimates are the best-fit values also used in the deposited code
  (`Value_BestEst`); the intervals summarise the 25 accepted multistart
  parameter sets. The point estimates are used.
- **Parameter-file label.** In the deposited `rateconstants_*_BEST.txt` the
  sixth row (26.513) is labelled `k50P`; the code reads it positionally as the
  TE half-maximal constant K50, matching Table S2.
- **`abs()` guards.** The in vitro code wraps the state vector and `1 - GRP`
  in `abs()`; both files keep `abs(1 - GRP)` (and the in vitro file
  `abs(LSD1TOTAL - LSD1B)`), which is inactive because GRP never exceeds 1
  and bound LSD1 never exceeds the total, but prevents a round-off negative
  base being raised to the non-integer Hill powers.
- **Alternative PK model not extracted.** The nonlinear-clearance PK model
  (Eq. 21; ka 0.15 1/h, k12 2.02 1/h, k21 0.05 1/h, V 80.3 mL/kg,
  KM 27.2 ng/mL, Vmax 68.7 ng/(mL*h)) was a comparator the authors rejected in
  favour of nonlinear absorption (Figure S3); it is not part of the final
  model.
- **No variability.** The parameters were estimated by least squares on mean
  data, with no between-animal variability or residual-error model; both files
  are deterministic. The paper's Figure S2 illustrates between-mouse spread by
  varying kP between 0.0022 and 0.0048 1/h, which users can reproduce by
  overriding `lkp`.
- **Observed data.** The observed points in the figures are the means in the
  authors' deposited `experimental_data.xlsx`.
- No erratum or correction notice for this article was found (checked
  2026-09-25).
