---
title: "Automated NODE-LASSO model development (Braem 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Automated NODE-LASSO model development (Braem 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Braem 2026 is a methods paper. It proposes an automated pharmacometric
model-development workflow in which a **low-dimensional neural ODE (NODE)** is
fitted to the data first, its learned right-hand side is read out as
derivative-versus-state data, and **LASSO regression** then selects an explicit,
mechanism-based set of functions that can replace the neural network. The paper
demonstrates it on three examples: neonatal weight, simulated bi-exponential PK,
and warfarin PK/PD.

Three of those models are packaged here.

```{r models}
model_names <- c(
  "Bram_2026_warfarin_node",
  "Bram_2026_biexponential",
  "Bram_2026_biexponential_2cmt"
)
uis <- lapply(model_names, function(n) rxode2::rxode(readModelDb(n)))
names(uis) <- model_names
```

* Citation: `r uis[["Bram_2026_warfarin_node"]]$reference`
* Article: <https://doi.org/10.1002/psp4.70285>
* Supporting Information (Tables S1/S2, derivations, Figures S1-S3) and the
  authored **Data S2 code deposit** are open access alongside the article.

## What is packaged, and what is not

Braem 2026 prints **no parameter table anywhere in the article text** -- Table 1
reports only mean absolute relative errors. Every value used here comes from the
Supporting Information or from the Data S2 code deposit, and the three models
below are exactly those for which estimates exist.

```{r coverage, echo = FALSE}
tibble::tribble(
  ~"Paper model", ~"Packaged", ~"Where the numbers are",
  "Warfarin NODE (Eq 16)", "yes -- `Bram_2026_warfarin_node`",
  "Data S2, Monolix run `Warf_node_mlx_file_ind/populationParameters.txt`",
  "Bi-exponential proposed one-state model (Eq 20)", "yes -- `Bram_2026_biexponential`",
  "Table S2 (`k`, `p`) plus an exact inversion for the other two",
  "Bi-exponential reference two-compartment model (Eq 14)", "yes -- `Bram_2026_biexponential_2cmt`",
  "Table S2 macro-parameters, inverted to micro-constants",
  "Neonatal weight model (Eq 18)", "no",
  "`kw, kt, q, p, W0` are stated to have been estimated; values printed nowhere",
  "Warfarin LASSO-proposed model (Eq 24)", "no",
  "parameters stated to have been estimated; values printed nowhere",
  "Warfarin reference Imax model (Eq 25)", "no",
  "only its BIC (2166) is reported"
) |>
  knitr::kable(caption = "Coverage of Braem 2026 in nlmixr2lib.")
```

The two omitted warfarin/weight models are excluded on the initial-versus-final
rule, not for want of looking. Figure 4 of the article *does* carry readable
numbers, and the Supporting Information even ships one derivative vector
verbatim -- but the Figure 4 caption states that the red dashed line is "the
through LASSO identified function", i.e. the step-3 output that becomes the
*starting values* for the step-4 fit. Those are initial estimates, and shipping
them as final would misrepresent the paper. This is a general shape worth
naming: **a derivative-versus-state plot in a NODE-distillation paper is an
initial-estimate figure by construction.**

# Population

The **warfarin** example uses the warfarin PK/PD dataset distributed as an
example with Monolix and nlmixr, originating from O'Reilly and Aggeler (1968):
32 adults given a single oral dose, with a median of 6 (IQR 6-10) plasma
warfarin concentrations and 7 (IQR 7-8) prothrombin complex activity (PCA)
observations each, over 144 h (Braem 2026 Section 2.3.3). The deposited
`warfarin_pkpd_data.csv` shows doses of 60-153 mg (mean 105 mg). Braem 2026
gives no baseline demographic table, and the deposited file carries no age,
weight, sex or race column.

The **bi-exponential** example is not a real study at all: Braem 2026
Section 2.3.2 simulated 50 subjects from the two-compartment model of
Equation 14, with 7 concentration measurements over 24 h, as a sensitivity
analysis asking whether the workflow recovers a usable structure for data that
arose from a model with a latent state.

```{r population}
str(uis[["Bram_2026_warfarin_node"]]$population)
```

# Source trace

Per-parameter origin is recorded as an in-file comment beside each `ini()`
entry. Collected here for review:

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~"Equation / parameter", ~"Value", ~"Source location",
  "PK structure: 1-cmt oral with lag", "n/a", "Eq 16, first three lines (Monolix `pkmodel(Tlag, ka, V, Cl)`)",
  "PD structure: indirect response, NN production", "n/a", "Eq 16, fourth line; expanded in Data S2 `Warf_node_converted.txt`",
  "Softplus activation, beta", "20", "Supporting Information Eq (2); pmxNODE default",
  "`ltlag`, `lka`, `lvc`, `lcl`", "0.8083 h, 1.241 /h, 8.003 L, 0.1308 L/h", "Data S2 `Warf_node_mlx_file_ind/populationParameters.txt`",
  "`lrbase`, `lkout`", "95.15 %, 0.04075 /h", "same file (`R0_pop`, `kout_pop`)",
  "`nn_w1_rc_*`, `nn_b1_rc_*`, `nn_w2_rc_*`", "15 values", "same file (`Wrc_1*`, `brc_1*`, `Wrc_2*`)",
  "21 IIV variances", "omega squared", "same file (`omega_*`); Monolix reports SD, nlmixr2 takes variance",
  "Residual error, both outputs", "`combined1(a, b)`", "Data S2 `Warf_node_mlx_file_ind.mlxtran`, DEFINITION block",
  "Proposed one-state model structure", "n/a", "Eq 20; multiple-dose `tad` form in Supporting Information",
  "`lkel` (paper `k` / `r`), `lkdist` (paper `p`)", "0.12 /h, 0.46 /h", "Table S2",
  "`lwdist` (paper `w` / `q`), `lvc`", "0.1717 /h, 1/10.1", "inverted from Table S2 `A'` = 5.1, `B'` = 5.0 (derived below)",
  "Two-compartment reference structure", "n/a", "Eq 14",
  "`lcl`, `lvc`, `lq`, `lvp`", "macro-to-micro inversion", "Table S2 A = 5.1, alpha = 0.46, B = 5.0, beta = 0.12 (derived below)"
) |>
  knitr::kable(caption = "Source trace for all three packaged models.")
```

# Part 1 -- the warfarin neural ODE

The PD production rate of this indirect-response model is a one-hidden-layer,
five-neuron softplus network of the plasma concentration:

$$
\frac{dPCA}{dt} = \sum_{j=1}^{5} w^{(2)}_j \cdot
  \frac{1}{\beta}\log\!\left(1 + e^{\,\beta \left(-\left(w^{(1)}_j\right)^2 C_c\right) + b^{(1)}_j}\right)
  - k_{out}\, PCA .
$$

## Gate 1 -- the network reproduces Figure 4E

Braem 2026 Figure 4E plots this network's output against concentration, running
from about 2.0 at the left-hand edge of the plotted range down to approximately
0 by 15 mg/L. Evaluating the packaged parameters directly is a check on the
transcription **and** on the operator precedence inside the softplus, which the
deposited code writes as `beta * (-(w^2) * Cc) + b` rather than
`beta * (-(w^2) * Cc + b)`.

```{r gate-nn}
ini_w <- uis[["Bram_2026_warfarin_node"]]$theta
nn_eval <- function(cc) {
  h <- vapply(1:5, function(j) {
    w1 <- ini_w[[paste0("nn_w1_rc_", j)]]
    b1 <- ini_w[[paste0("nn_b1_rc_", j)]]
    log(1 + exp(20 * (-(w1^2) * cc) + b1)) / 20
  }, numeric(1))
  w2 <- vapply(1:5, function(j) ini_w[[paste0("nn_w2_rc_", j)]], numeric(1))
  sum(w2 * h)
}
nn_curve <- tibble(Cc = seq(0, 20, by = 0.1)) |>
  mutate(production = vapply(Cc, nn_eval, numeric(1)))

edge_left  <- nn_eval(0.5)
edge_right <- nn_eval(15)
c(`NN at Cc = 0.5` = edge_left, `NN at Cc = 15` = edge_right)

# Deterministic evaluation of published constants: a tight bound is correct
# here, there is no cohort and no RNG involved.
stopifnot(
  abs(edge_left - 2.0) < 0.1,   # Figure 4E left-hand edge reads 2.0
  edge_right < 0.1,             # Figure 4E has decayed to ~0 by 15 mg/L
  all(diff(nn_curve$production) <= 0)  # softplus with negated squared weights is monotone
)
```

```{r figure-4e, echo = FALSE}
ggplot(nn_curve, aes(Cc, production)) +
  geom_line(linewidth = 1) +
  labs(x = "Warfarin concentration (mg/L)",
       y = "PCA production rate (% of normal per h)",
       title = "Figure 4E -- learned production term",
       caption = "Replicates Figure 4E of Braem 2026.") +
  theme_bw()
```

## Gate 2 -- the packaged model reproduces Monolix's own population predictions

The Data S2 deposit contains the fitted run's `predictions_ycp.txt` and
`predictions_ypca.txt`, which hold Monolix's population predictions (`popPred`,
i.e. all etas at zero) at every observation time of every subject. Because the
population prediction depends on nothing but the subject's dose, solving the
packaged model with `zeroRe()` at that dose must reproduce them. This is a
**zero-parameter gate**: it simultaneously confirms all 21 estimates, the ODE
structure, the softplus precedence, and the `PCA(0) = rbase` initial condition.

Three subjects spanning the dose range are transcribed below.

```{r gate-poppred}
published_pred <- tibble::tribble(
  ~id, ~amt, ~time, ~output, ~popPred,
  # --- subject 1, 100 mg (Data S2 predictions_ypca.txt / predictions_ycp.txt)
  1L, 100,  24,   "PCA", 37.0193,
  1L, 100,  36,   "PCA", 23.9270,
  1L, 100,  48,   "PCA", 18.9423,
  1L, 100,  72,   "PCA", 21.0226,
  1L, 100,  96,   "PCA", 27.9879,
  1L, 100, 120,   "PCA", 34.7860,
  1L, 100, 144,   "PCA", 40.1675,
  1L, 100,   1,   "Cc",   2.6408,
  1L, 100,   2,   "Cc",   9.5320,
  1L, 100,   3,   "Cc",  11.3820,
  1L, 100,   6,   "Cc",  11.6114,
  1L, 100,   9,   "Cc",  11.0746,
  1L, 100,  12,   "Cc",  10.5451,
  1L, 100,  24,   "Cc",   8.6672,
  1L, 100,  36,   "Cc",   7.1238,
  1L, 100,  48,   "Cc",   5.8551,
  1L, 100,  72,   "Cc",   3.9554,
  # --- subject 5, 60 mg
  5L,  60,   0,   "PCA", 95.1495,
  5L,  60,  24,   "PCA", 44.9263,
  5L,  60,  36,   "PCA", 36.1485,
  5L,  60,  48,   "PCA", 32.8811,
  5L,  60,  72,   "PCA", 33.8858,
  5L,  60,  96,   "PCA", 37.9848,
  5L,  60, 120,   "PCA", 42.0326,
  5L,  60, 144,   "PCA", 45.2464,
  5L,  60,   3,   "Cc",   6.8292,
  5L,  60,   6,   "Cc",   6.9669,
  5L,  60,  12,   "Cc",   6.3271,
  5L,  60,  24,   "Cc",   5.2003,
  5L,  60,  48,   "Cc",   3.5131,
  5L,  60, 120,   "Cc",   1.0831,
  # --- subject 12, 123 mg
  12L, 123,   0,  "PCA", 95.1495,
  12L, 123,  24,  "PCA", 36.9130,
  12L, 123,  36,  "PCA", 22.7836,
  12L, 123,  48,  "PCA", 15.0629,
  12L, 123,  72,  "PCA", 15.1787,
  12L, 123,  96,  "PCA", 22.8206,
  12L, 123, 120,  "PCA", 30.8389,
  12L, 123, 144,  "PCA", 37.3307,
  12L, 123, 1.5,  "Cc",   8.7979,
  12L, 123,   3,  "Cc",  13.9998,
  12L, 123,   6,  "Cc",  14.2821,
  12L, 123,  12,  "Cc",  12.9705,
  12L, 123,  24,  "Cc",  10.6607,
  12L, 123,  48,  "Cc",   7.2018,
  12L, 123, 120,  "Cc",   2.2203
)

warf_typical <- rxode2::zeroRe(uis[["Bram_2026_warfarin_node"]])

solve_one <- function(subject_id, dose, times) {
  ev <- rxode2::et(amt = dose, cmt = "depot") |>
    rxode2::et(sort(unique(c(0, times))))
  ev <- as.data.frame(ev)
  # Two `~` endpoints: put dvid on every observation row. rxSolve returns
  # BOTH Cc and PCA as columns regardless, so one grid suffices.
  ev$dvid <- ifelse(ev$evid == 0, 1L, NA_integer_)
  s <- rxode2::rxSolve(warf_typical, ev, returnType = "data.frame")
  tibble(id = subject_id, time = s$time, Cc = s$Cc, PCA = s$PCA)
}

simulated_pred <- published_pred |>
  group_by(id) |>
  group_modify(function(d, k) {
    solve_one(k$id, d$amt[1], d$time) |> select(-id)
  }) |>
  ungroup()

poppred_check <- published_pred |>
  left_join(simulated_pred, by = c("id", "time")) |>
  mutate(simulated = if_else(output == "PCA", PCA, Cc),
         pct_diff  = 100 * (simulated - popPred) / popPred) |>
  filter(popPred > 0)   # the pre-absorption Cc rows are exactly 0 in both

max_poppred_diff <- max(abs(poppred_check$pct_diff))
max_poppred_diff

# Deterministic (zeroRe, fixed doses, no RNG): a tight bound is correct.
# Realised 0.213% here; the residual is Monolix's own reporting precision and
# integrator tolerance. A single mis-transcribed weight, bias or rate moves
# this by whole percent or more.
stopifnot(max_poppred_diff < 1)
```

```{r poppred-table, echo = FALSE}
poppred_check |>
  group_by(id, output) |>
  summarise(n = n(),
            `max |% diff|` = max(abs(pct_diff)),
            .groups = "drop") |>
  rename("Subject" = id, "Output" = output, "Observations" = n) |>
  knitr::kable(digits = 3,
               caption = "Packaged model vs Monolix popPred from the Data S2 deposit.")
```

```{r poppred-figure, echo = FALSE}
poppred_check |>
  mutate(panel = if_else(output == "PCA",
                         "PCA (% of normal)", "Warfarin (mg/L)")) |>
  ggplot(aes(popPred, simulated)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey50") +
  geom_point(aes(colour = factor(id)), size = 2) +
  facet_wrap(~panel, scales = "free") +
  labs(x = "Monolix popPred (Data S2)", y = "nlmixr2lib model",
       colour = "Subject",
       title = "Zero-parameter replication of the deposited fit") +
  theme_bw()
```

## The baseline is deliberately not a steady state

Because the network replaces the production term outright, nothing constrains it
to balance `kout * PCA0` at zero concentration. It does not:

```{r baseline-drift}
kin_implied  <- nn_eval(0)
loss_at_base <- ini_w[["lkout"]] |> exp() * exp(ini_w[["lrbase"]])
c(`network production at Cc = 0` = kin_implied,
  `kout * PCA0`                  = loss_at_base)

stopifnot(kin_implied < loss_at_base)   # deterministic; both are fixed constants
```

So PCA declines from baseline even with no drug present. That asymmetry is
precisely what Braem 2026 repairs in the LASSO-distilled model of Equation 24,
by rearranging the identified functions and introducing `kin = k + q` "to ensure
steady-state behavior when no drug is given". It is worth seeing in the packaged
NODE, because it is the concrete cost of a black-box right-hand side that the
paper's whole workflow exists to convert into an interpretable one.

## Simulated cohort and PKNCA

```{r cohort}
# rxSetSeed() fixes rxode2's stream per solver thread, not across thread
# counts, so this cohort differs between machines. Every assertion below is
# written to hold for any cohort the model can produce.
rxode2::rxSetSeed(20260903)
n_sub <- 100L   # well under the 200-per-arm cap

obs_times <- c(seq(0, 12, by = 0.5), seq(14, 144, by = 2))
cohort_ev <- rxode2::et(amt = 100, cmt = "depot") |> rxode2::et(obs_times)
cohort_ev <- as.data.frame(cohort_ev)
cohort_ev$dvid <- ifelse(cohort_ev$evid == 0, 1L, NA_integer_)
cohort_ev$treatment <- "100 mg single oral dose"

sim <- rxode2::rxSolve(uis[["Bram_2026_warfarin_node"]], cohort_ev,
                       nSub = n_sub, keep = "treatment",
                       returnType = "data.frame")

# The event table holds a single subject replicated by `nSub`, so rxSolve
# returns the subject key as `sim.id` and omits `id` entirely. Restore it
# before anything downstream groups on it.
if (is.null(sim$id)) sim$id <- sim$sim.id
stopifnot(dplyr::n_distinct(sim$id) == n_sub)
```

```{r figure-3c, echo = FALSE}
sim |>
  select(id, time, Cc, PCA) |>
  pivot_longer(c(Cc, PCA), names_to = "output", values_to = "value") |>
  mutate(panel = if_else(output == "PCA",
                         "PCA (% of normal)", "Warfarin (mg/L)")) |>
  group_by(panel, time) |>
  summarise(Q05 = quantile(value, 0.05), Q50 = median(value),
            Q95 = quantile(value, 0.95), .groups = "drop") |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~panel, scales = "free_y") +
  labs(x = "Time (h)", y = NULL,
       title = "Simulated cohort, 100 mg single oral dose",
       caption = "Compare Figure 3C of Braem 2026 (observed data with NODE population fit).") +
  theme_bw()
```

The paper reports no NCA values for warfarin, so the NCA below is checked
against the quantity the model itself determines: for a linear
one-compartment oral model with complete bioavailability,
`AUC(0-inf) = Dose / CL`.

```{r pknca-warfarin}
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)

sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df  <- cohort_ev |>
  filter(evid == 1) |>
  select(time, amt, treatment) |>
  tidyr::crossing(id = unique(sim_nca$id))
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  conc_obj, dose_obj,
  intervals = data.frame(start = 0, end = Inf,
                         cmax = TRUE, tmax = TRUE,
                         aucinf.obs = TRUE, half.life = TRUE)
))

nca_wide <- as.data.frame(nca_res) |>
  select(id, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

nca_summary <- nca_wide |>
  summarise(across(c(cmax, tmax, aucinf.obs, half.life),
                   list(median = ~median(.x, na.rm = TRUE)),
                   .names = "{.col}"))

cl_typical  <- exp(ini_w[["lcl"]])
auc_expected <- 100 / cl_typical

tibble::tibble(
  `NCA parameter` = c("Cmax (mg/L)", "Tmax (h)", "AUC0-inf (mg*h/L)", "t1/2 (h)"),
  `Simulated median` = c(nca_summary$cmax, nca_summary$tmax,
                         nca_summary$aucinf.obs, nca_summary$half.life),
  `Model-implied reference` = c(NA, NA, auc_expected, log(2) / (cl_typical / exp(ini_w[["lvc"]])))
) |>
  knitr::kable(digits = 2,
               caption = "Simulated NCA vs the values the packaged PK parameters imply.")
```

```{r pknca-warfarin-gate}
auc_pct <- 100 * (nca_summary$aucinf.obs - auc_expected) / auc_expected
auc_pct

# Cohort-derived, so bound the magnitude with headroom rather than pinning a
# value from one draw. The median of a log-normal AUC is Dose/CL exactly, so
# the residual is not a distributional offset: it is `aucinf.obs` tail
# extrapolation. Observations stop at 144 h, only about 3.4 terminal
# half-lives, so a large extrapolated fraction is fitted from a log-linear
# slope that the simulated cohort reads slightly long (44.7 h vs the
# model-implied 42.4 h) -- which inflates AUC. The residual is also strongly
# cohort-dependent, because rxode2 partitions its RNG streams per solver
# thread: realised +4.73% at 16 threads and -1.59% at 2. Do not tighten to
# either. A mis-transcribed dose, volume or clearance moves this by tens of
# percent, so 15 still goes red for the failures that matter.
stopifnot(abs(auc_pct) < 15)
```

# Part 2 -- the automatically-proposed bi-exponential model

The LASSO step selected a linear function of amount plus an exponential function
of time, giving (Equation 20)

$$
\frac{dA_C}{dt} = -k_{el} A_C - D \cdot w_{dist}\, e^{-k_{dist} t},
\qquad A_C(0) = D, \qquad C_C = A_C / V_c .
$$

There is no peripheral compartment. The paper's claim -- and the reason the
example is interesting -- is that this single-state model has an explicit
solution *identical* to the central compartment of the two-compartment model
that generated the data.

## Recovering the parameters from Table S2

Braem 2026 never prints the estimates of Equation 20. Table S2 gives the
macro-parameters of its explicit solution (Equation 21),
`Cc(t) = A' exp(-kdist t) + B' exp(-kel t)`, together with `kel` and `kdist`
themselves. The paper's own definitions of `A'` and `B'` then invert exactly.

```{r invert}
Ap <- 5.1; kdist <- 0.46; Bp <- 5.0; kel <- 0.12   # Table S2

# A' = -(D/Vc) * wdist / (kel - kdist);  B' = (D/Vc) * (kel - kdist + wdist) / (kel - kdist)
# Summing telescopes the bracket:  A' + B' = D/Vc.
d_over_v <- Ap + Bp
wdist    <- Ap * (kel - kdist) / -d_over_v

c(`D/Vc` = d_over_v, wdist = wdist)

# Round-trip the inversion back through the paper's own formulas.
stopifnot(
  abs(-d_over_v * wdist / (kel - kdist) - Ap) < 1e-9,
  abs(d_over_v * (kel - kdist + wdist) / (kel - kdist) - Bp) < 1e-9
)

# And the standard macro-to-micro inversion for the two-compartment reference.
A <- 5.1; alpha <- 0.46; B <- 5.0; beta <- 0.12    # Table S2
k21 <- (A * beta + B * alpha) / (A + B)
k10 <- alpha * beta / k21
k12 <- alpha + beta - k10 - k21
vc2 <- 1 / (A + B)

stopifnot(
  abs((k10 + k12 + k21) - (alpha + beta)) < 1e-12,
  abs(k10 * k21 - alpha * beta) < 1e-12
)
c(vc = vc2, cl = k10 * vc2, q = k12 * vc2, vp = k12 * vc2 / k21)
```

The fit identifies the model only up to the ratio `D/Vc = 10.1`; the dose was
never reported. Both packaged models therefore fix a nominal **unit dose**, so
`vc = 1/10.1`, and a dose of 1 arbitrary unit reproduces Table S2 exactly. The
model is linear, so any other dose scales the profile exactly.

## Gates 3-5 -- both models against their closed forms, and against each other

```{r gate-biexp}
biexp_ev <- rxode2::et(amt = 1, cmt = "central") |>
  rxode2::et(seq(0, 24, by = 0.1))

sim_prop <- rxode2::rxSolve(uis[["Bram_2026_biexponential"]], biexp_ev,
                            returnType = "data.frame")
sim_2cmt <- rxode2::rxSolve(uis[["Bram_2026_biexponential_2cmt"]], biexp_ev,
                            returnType = "data.frame")

closed_form <- 5.1 * exp(-0.46 * sim_prop$time) + 5.0 * exp(-0.12 * sim_prop$time)

gates <- c(
  `proposed vs Eq 21` = max(abs(sim_prop$Cc - closed_form) / closed_form),
  `2-cmt vs Eq 22`    = max(abs(sim_2cmt$Cc - closed_form) / closed_form),
  `proposed vs 2-cmt` = max(abs(sim_prop$Cc - sim_2cmt$Cc) / sim_2cmt$Cc)
)
gates

# These are deterministic solves of published constants against a published
# closed form -- no cohort, no RNG -- so a tight bound is correct and is what
# catches a regression. Realised ~5e-07.
stopifnot(all(gates < 1e-4))
```

That third number is the paper's central structural claim, reproduced: a model
with **no** peripheral compartment and a model **with** one give the same
central-compartment concentration to seven significant figures.

```{r figure-biexp, echo = FALSE}
bind_rows(
  tibble(time = sim_prop$time, Cc = sim_prop$Cc, model = "Proposed one-state (Eq 20)"),
  tibble(time = sim_2cmt$time, Cc = sim_2cmt$Cc, model = "Two-compartment reference (Eq 14)")
) |>
  ggplot(aes(time, Cc, colour = model, linetype = model)) +
  geom_line(linewidth = 1) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Concentration (arbitrary units)",
       colour = NULL, linetype = NULL,
       title = "Identical central-compartment kinetics from different structures",
       caption = "Braem 2026 Equations 20-22 and Table S2; nominal unit dose.") +
  theme_bw() +
  theme(legend.position = "bottom")
```

## PKNCA on the pair

```{r pknca-biexp}
biexp_conc <- bind_rows(
  tibble(id = 1L, time = sim_prop$time, Cc = sim_prop$Cc,
         model = "Proposed one-state"),
  tibble(id = 1L, time = sim_2cmt$time, Cc = sim_2cmt$Cc,
         model = "Two-compartment")
) |>
  filter(!is.na(Cc)) |>
  arrange(model, id, time)

biexp_dose <- tibble(id = 1L, time = 0, amt = 1,
                     model = unique(biexp_conc$model))

biexp_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(biexp_conc, Cc ~ time | model + id),
  PKNCA::PKNCAdose(biexp_dose, amt ~ time | model + id),
  intervals = data.frame(start = 0, end = Inf,
                         cmax = TRUE, tmax = TRUE,
                         aucinf.obs = TRUE, half.life = TRUE)
))

# Table S2 fixes AUC(0-inf) analytically: A/alpha + B/beta.
auc_published <- A / alpha + B / beta

biexp_tab <- as.data.frame(biexp_nca) |>
  select(model, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

biexp_tab |>
  mutate(`AUC from Table S2` = auc_published,
         `AUC % diff` = 100 * (aucinf.obs - auc_published) / auc_published) |>
  rename("Model" = model, "Cmax" = cmax, "Tmax (h)" = tmax,
         "AUC0-inf" = aucinf.obs, "t1/2 (h)" = half.life) |>
  knitr::kable(digits = 4,
               caption = "NCA of both models vs the AUC that Table S2 determines analytically.")

# Deterministic solve + trapezoidal NCA on a 0.1 h grid; realised < 0.05%.
stopifnot(
  max(abs(100 * (biexp_tab$aucinf.obs - auc_published) / auc_published)) < 1,
  # Both models must give the same Cmax: it is the same profile.
  diff(range(biexp_tab$cmax)) < 1e-4
)
```

## Gate 6 -- the documented multiple-dose limitation

Braem 2026 Figure S1 shows that the proposed one-state model is structurally
misspecified when a dose arrives *before* the previous distribution phase has
finished, because the model carries no memory of earlier doses. That is a
property of the model, not a defect of this encoding, and it is worth
demonstrating so a user does not walk into it.

```{r gate-multidose}
compare_regimen <- function(ii, n_doses) {
  ev <- rxode2::et(amt = 1, cmt = "central", ii = ii, addl = n_doses - 1L) |>
    rxode2::et(seq(0, ii * n_doses, by = 0.1))
  p <- rxode2::rxSolve(uis[["Bram_2026_biexponential"]], ev, returnType = "data.frame")
  r <- rxode2::rxSolve(uis[["Bram_2026_biexponential_2cmt"]], ev, returnType = "data.frame")
  max(abs(p$Cc - r$Cc) / pmax(r$Cc, 1e-8))
}

multidose <- c(
  `q24h (dose after distribution phase)` = compare_regimen(24, 4L),
  `q2h  (dose within distribution phase)` = compare_regimen(2, 12L)
)
multidose

# Deterministic. The q24h agreement is the paper's stated validity condition;
# the q2h disagreement is the Figure S1 misspecification, and asserting that it
# is LARGE is what stops a future refactor from silently "fixing" the model
# into something the paper did not publish.
stopifnot(
  multidose[[1]] < 0.05,
  multidose[[2]] > 0.20
)
```

# Assumptions and deviations

* **No parameter table exists in the article.** Every value in all three models
  comes from the Supporting Information (Table S2) or from the authored Data S2
  code deposit. This is recorded per parameter in the model files and in the
  source-trace table above.
* **Warfarin NODE parameters are non-paper-derived.** All 21 estimates and both
  residual-error models come from the deposited Monolix run
  `Warf_node_mlx_file_ind` (`populationParameters.txt` and the `.mlxtran`
  DEFINITION block). The deposit contains two runs; the one used here is the
  final one, with IIV on the network parameters, which is the fit whose MARE the
  article reports.
* **Monolix omega is a standard deviation; nlmixr2 takes a variance.** Every
  IIV entry in the warfarin model is the deposited omega squared, noted inline.
* **The additive residual SD on `Cc` collapsed to zero.** The deposit reports
  `acp = 2.22e-16`, the optimiser's lower bound. It is transcribed verbatim
  rather than dropped, so the `combined1` error model matches the deposit
  exactly; in practice the concentration residual is effectively proportional.
* **The bi-exponential parameters `wdist` and `Vc` are back-solved, not
  printed.** Table S2 gives `A'`, `B'`, `kel` and `kdist`; `wdist` and `D/Vc`
  follow exactly from the paper's own definitions of `A'` and `B'`, and the
  inversion is round-tripped in the vignette above.
* **The bi-exponential dose is unreported, so a nominal unit dose is fixed.**
  Both bi-exponential models are identified only up to `D/Vc = 10.1`. Setting
  `D = 1` gives `vc = 1/10.1`; the model is linear, so any other dose scales the
  profile exactly and nothing is invented beyond the published ratio. Units for
  dose and concentration are declared as placeholders (`dose_unit`,
  `conc_unit`) because the source data were simulated and carry none.
* **No IIV and no residual error for either bi-exponential model.** Braem 2026
  states that parameters were assumed log-normally distributed but reports no
  omega and no sigma for Equations 14 or 20. Nothing is invented; both models
  are typical-value only.
* **Two of the paper's models are not packaged.** The neonatal weight model
  (Equation 18) and the LASSO-distilled warfarin model (Equation 24) report no
  final estimates anywhere on disk. See the coverage table above for why the
  Figure 4 panels do not rescue them.
* **`podo()` and `tad()` need their compartment argument.** In
  `Bram_2026_biexponential`, the transient flux is written
  `podo(central) * wdist * exp(-kdist * tad(central))`. Verified on rxode2
  5.1.7: through the `rxUi` path that `readModelDb()` and `rxSolve()` use, the
  argument-less `podo()` / `tad()` silently evaluate to zero, which drops the
  term and degrades the model to a mono-exponential with no warning. Gate 3
  above exists to catch exactly that regression.
