---
title: "Meropenem (Braune 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Meropenem (Braune 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Braune_2018_meropenem"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1186/s13054-018-1940-1>
* PubMed Central: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5791175/>

Braune 2018 is a prospective single-centre population PK study of meropenem in
critically ill septic adults with acute kidney injury who were receiving
**sustained low-efficiency dialysis (SLED)**, a prolonged-intermittent renal
replacement modality. The clinically interesting feature -- and the reason the
model is worth packaging -- is that meropenem clearance is not a single number:
it switches between a low interdialytic value and a much higher intradialytic
value every time a SLED session starts and stops, and the interdialytic value
itself depends on how much urine the patient still makes.

# Population

`r ui$population$n_subjects` septic adults (`r 100 - ui$population$sex_female_pct`%
male) contributed `r ui$population$n_concentrations` serum meropenem
concentrations. Median age was 66 years [range 37-78], median weight 81 kg
[range 70-183], and median SOFA score on the first sampling day 11 [range 5-16].
All were on SLED for acute kidney injury; **the median residual diuresis was
0 mL/24h**, i.e. the median patient was anuric, with an interquartile range of
0-80 mL/24h and a maximum of 360 mL/24h (Braune 2018 Table 1). ICU mortality was
47%.

Meropenem was given as 0.5 g, 1 g or 2 g IV over 30 minutes 8-hourly at the
discretion of the treating physician. Sampling ran over three consecutive SLED
days: a trough 1 h before the infusion, then 10 min, 1 h, 2 h and 4 h after the
start of SLED, and at the end of the session. Post-SLED samples were **not**
collected, so any post-dialysis rebound is unobserved (Braune 2018 Limitations).

The same information is available programmatically via
`readModelDb("Braune_2018_meropenem")()$population`.

# Source trace

Every `ini()` entry carries an in-file comment naming its source location; the
table below collects them for review. All parameter values come from **Table 2**
("Parameter estimates for meropenem from the final covariate two-compartment
population pharmacokinetic model"), whose columns are Mean / Standard deviation /
Coefficient of variation (%) / Median. The typical values below are the **Mean**
column -- the column the Discussion quotes throughout.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl_nonren` (CLD) | `log(2.6)` L/h | Table 2, row `CLD`: mean 2.6, SD 1.2, CV 44.9%, median 2.3 |
| `lcl_renal` (CLN) | `log(1.5)` L/h per 100 mL/24h | Table 2, row `CLN`: mean 1.5, SD 2.1, CV 134.7%, median 0.7 |
| `lcl_crrt` (CLSLED) | `log(7.9)` L/h | Table 2, row `CLSLED`: mean 7.9, SD 4.2, CV 53.6%, median 6.8 |
| `lvc` (Vc) | `log(8.1)` L | Table 2, row `Vc`: mean 8.1, SD 7.1, CV 87.9%, median 4.9 |
| `lk12` (Kcp) | `log(10.3)` 1/h | Table 2, row `KCP`: mean 10.3, SD 8.8, CV 85.4%, median 7.9 |
| `lk21` (Kpc) | `log(1.8)` 1/h | Table 2, row `KPC`: mean 1.8, SD 1.9, CV 104.4%, median 1.2 |
| `etalcl_nonren` ... `etalk21` | `log(CV^2 + 1)` | Table 2, "Coefficient of variation (%)" column |
| `addSd` | `fixed(0)` | Methods / Results state an additive (lambda) error model; the magnitude is reported nowhere |
| `cl <- cl_nonren + cl_renal + RRT_CRRT_ACTIVE * cl_crrt` | n/a | Results, "Pharmacokinetic model building": `TVCL = CLSLED * (SLED) + CLNS` |
| `cl_renal <- exp(lcl_renal + eta) * URINE_VOL_24H / 100` | n/a | Same equation block: `TVCLNS = CLD + (CLN * RD / 100)` |
| Two-compartment ODEs in Kcp / Kpc | n/a | Table 2 parameterisation; Abbreviations list defines KCP / KPC |
| SLED indicator semantics | 1 on / 0 off | Results: "The value for term SLED is 1 when SLED is on, whereas it is 0 when SLED is off" |

The two model equations are printed in the article as a single typeset image.
The PDF text layer silently drops the multiplication signs from that image
(`TVCL = CLSLED (SLED) + CLNS`), which reads ambiguously; the publisher's
figure-file bundle for PMC5791175 contains the image itself
(`13054_2018_1940_Article_Equa.gif`), which renders unambiguously as
`TVCL = CL_SLED * (SLED) + CL_NS`. The SLED arm is therefore **additive** on top
of the interdialytic arms, not a replacement for them. See "Assumptions and
deviations" for the one Discussion sentence that reads loosely against this.

# Structural verification

These checks compare the packaged model against quantities that follow
algebraically from the published parameters. They are **deterministic** -- typical
values with the random effects zeroed -- so the difference between the two sides
is pure numerical error and a tight bound is the correct gate.

```{r helpers}
mod <- readModelDb("Braune_2018_meropenem")
mod_typical <- rxode2::zeroRe(rxode2::rxode(mod))

# Published parameters (Braune 2018 Table 2, Mean column).
p <- list(cld = 2.6, cln = 1.5, clsled = 7.9, vc = 8.1, kcp = 10.3, kpc = 1.8)

# The paper's printed clearance equation, evaluated directly.
published_cl <- function(rd, sled) p$clsled * sled + p$cld + p$cln * rd / 100

# Build an event table. `rate` (not `dur`) sets the infusion duration explicitly.
# RRT_CRRT_ACTIVE is time-varying, so it is supplied on EVERY record and the
# solve below uses covsInterpolation = "locf" -- linear interpolation would ramp
# a binary on/off gate across the transition instead of switching it.
make_events <- function(n, dose_mg, ii = 8, n_doses = 1L, rd = 0,
                        sled_start = NA_real_, sled_dur = 5, infusion_h = 0.5,
                        tmax = 48, dt = 0.1, id_offset = 0L) {
  dose_times <- seq(0, by = ii, length.out = n_doses)
  sled_edges <- if (is.na(sled_start)) numeric(0) else c(sled_start, sled_start + sled_dur)
  obs_times <- sort(unique(c(
    seq(0, tmax, by = dt),
    # resolve the fast distribution phase (alpha half-life ~3 min) after each dose
    as.vector(outer(dose_times, c(seq(0, infusion_h, by = 0.02), seq(0.6, 2, by = 0.05)), "+")),
    dose_times, dose_times + infusion_h, sled_edges
  )))
  obs_times <- obs_times[obs_times >= 0 & obs_times <= tmax]
  one <- dplyr::bind_rows(
    dplyr::tibble(time = dose_times, evid = 1L, amt = dose_mg,
                  rate = dose_mg / infusion_h, cmt = "central"),
    dplyr::tibble(time = obs_times, evid = 0L, amt = NA_real_,
                  rate = NA_real_, cmt = "central")
  ) |>
    dplyr::arrange(.data$time, dplyr::desc(.data$evid))
  tidyr::expand_grid(id = id_offset + seq_len(n), one) |>
    dplyr::mutate(
      URINE_VOL_24H = rd,
      RRT_CRRT_ACTIVE =
        if (is.na(sled_start)) 0 else
          as.numeric(.data$time >= sled_start & .data$time < sled_start + sled_dur)
    )
}

solve_typical <- function(events, ...) {
  rxode2::rxSolve(mod_typical, events, covsInterpolation = "locf",
                  returnType = "data.frame", ...)
}
```

## 1. The published clearance equation is reproduced exactly

For a linear model given intravenously, `CL * AUC(0-Inf) == Dose` identically.
Recovering the paper's own `TVCL` from the simulated AUC therefore tests the
whole covariate equation -- both arms of the interdialytic clearance, the
`RD / 100` normalisation, and the additive SLED gate -- in one number.

```{r gate-clearance}
grid_cl <- tidyr::expand_grid(rd = c(0, 100, 300), sled = c(0, 1))

cl_check <- grid_cl |>
  rowwise() |>
  mutate(
    sim = list({
      # SLED held constant for the whole profile so a single CL applies.
      ev <- make_events(1, dose_mg = 1000, rd = .data$rd, tmax = 240, dt = 0.05,
                        sled_start = if (.data$sled == 1) 0 else NA_real_,
                        sled_dur = 240)
      solve_typical(ev)
    }),
    cl_published = published_cl(.data$rd, .data$sled),
    cl_recovered = {
      o <- sim[!is.na(sim$Cc) & sim$time > 0, ]
      auc <- sum(diff(o$time) * (head(o$Cc, -1) + tail(o$Cc, -1)) / 2)
      lam <- -stats::coef(stats::lm(log(Cc) ~ time, data = tail(o, 400)))[[2]]
      1000 / (auc + tail(o$Cc, 1) / lam)
    },
    pct = 100 * (cl_recovered / cl_published - 1)
  ) |>
  ungroup() |>
  select(-sim)

cl_check |>
  mutate(across(c(cl_published, cl_recovered), \(x) round(x, 4)), pct = round(pct, 4)) |>
  dplyr::rename(
    "Residual diuresis (mL/24h)" = rd, "SLED running" = sled,
    "Published CL (L/h)" = cl_published, "Recovered Dose/AUCinf (L/h)" = cl_recovered,
    "Difference (%)" = pct
  ) |>
  knitr::kable(caption = paste(
    "Clearance recovered from the simulated AUC versus the paper's printed",
    "equation TVCL = CLSLED * SLED + CLD + CLN * RD / 100."
  ))

# Deterministic identity: the only error is trapezoidal/extrapolation error on a
# 0.05 h grid. Realised max |difference| was 0.02%.
stopifnot(max(abs(cl_check$pct)) < 0.5)
```

## 2. Two-compartment disposition is real, not collapsed to one compartment

This gate exists because of a specific and entirely silent rxode2 failure mode.
A model that defines variables named `cl` and `vc` but no `q` / `vp` pair can be
matched against rxode2's **one-compartment** analytic linear-compartment kernel:
the explicit `d/dt()` right-hand sides are then discarded, `peripheral1`
disappears from the solve, and the returned profile is mono-exponential decaying
at exactly `cl / vc`. Nothing flags it -- `ui$linCmt` reads empty, the ODE block
parses and prints correctly, `checkModelConventions()` is clean, and even the
clearance gate above still passes, because `CL * AUCinf == Dose` is blind to the
number of compartments.

The packaged model therefore also derives `vp <- vc * k12 / k21` and
`q <- k12 * vc`, which restores a consistent two-compartment specification. The
check below pins the solved profile to the closed-form biexponential so a
regression cannot pass silently.

```{r gate-biexponential}
# Closed form for an IV bolus into a two-compartment system parameterised by
# Vc, Kcp, Kpc and Kel (any PK text; e.g. Gibaldi & Perrier eq. 2.31).
biexp <- function(t, dose, vc, k12, k21, kel) {
  s <- kel + k12 + k21
  disc <- sqrt(s^2 - 4 * kel * k21)
  alpha <- (s + disc) / 2
  beta <- (s - disc) / 2
  (dose / vc) * (((alpha - k21) / (alpha - beta)) * exp(-alpha * t) +
                   ((k21 - beta) / (alpha - beta)) * exp(-beta * t))
}

ev_bolus <- make_events(1, dose_mg = 2000, rd = 0, tmax = 48, dt = 0.05,
                        infusion_h = 0) |>
  mutate(rate = ifelse(.data$evid == 1L, 0, .data$rate))
sim_bolus <- solve_typical(ev_bolus)

stopifnot("peripheral1" %in% names(sim_bolus)) # the state must survive the solve

obs_b <- sim_bolus[!is.na(sim_bolus$Cc) & sim_bolus$time > 0, ]
cf <- biexp(obs_b$time, 2000, p$vc, p$kcp, p$kpc, published_cl(0, 0) / p$vc)
rel_err <- max(abs(obs_b$Cc / cf - 1))

# Also confirm the profile is genuinely two-phase: a one-compartment collapse
# would decay at kel = CL/Vc from t = 0, which is 3.5-fold slower than the true
# alpha phase here.
early_slope <- -stats::coef(stats::lm(log(Cc) ~ time,
                                      data = obs_b[obs_b$time <= 0.25, ]))[[2]]
kel_1cmt <- published_cl(0, 0) / p$vc

cat(sprintf("max relative deviation from closed-form biexponential: %.3g\n", rel_err))
cat(sprintf("initial log-linear slope: %.3f 1/h  (kel = CL/Vc would be %.3f 1/h)\n",
            early_slope, kel_1cmt))

# Pure numerics against an exact algebraic solution -> a tight bound is correct.
stopifnot(rel_err < 1e-3)
# The alpha phase must be far faster than a one-compartment collapse would give.
stopifnot(early_slope > 5 * kel_1cmt)
```

## 3. Central volume and the SLED gate

```{r gate-vc-and-gate}
# C(0+) after a bolus is exactly Dose / Vc.
c0 <- sim_bolus$Cc[sim_bolus$time == 0 & !is.na(sim_bolus$Cc)][1]
cat(sprintf("C(0+) = %.4f mg/L; Dose/Vc = %.4f mg/L\n", c0, 2000 / p$vc))
stopifnot(abs(c0 / (2000 / p$vc) - 1) < 1e-4)

# The SLED gate must actually move the ODE states. A gate that reports a
# correct-looking `cl` column while the amounts decay identically in both states
# is the documented inert-gate defect; this compares the solved profiles.
ev_off <- make_events(1, 2000, rd = 0, tmax = 24, dt = 0.1)
ev_on <- make_events(1, 2000, rd = 0, tmax = 24, dt = 0.1,
                     sled_start = 0, sled_dur = 24)
s_off <- solve_typical(ev_off)
s_on <- solve_typical(ev_on)
auc_off <- sum(diff(s_off$time) * (head(s_off$Cc, -1) + tail(s_off$Cc, -1)) / 2, na.rm = TRUE)
auc_on <- sum(diff(s_on$time) * (head(s_on$Cc, -1) + tail(s_on$Cc, -1)) / 2, na.rm = TRUE)
cat(sprintf("AUC(0-24) SLED off = %.1f, SLED on = %.1f mg*h/L (ratio %.2f)\n",
            auc_off, auc_on, auc_off / auc_on))

# CL rises from 2.6 to 10.5 L/h, so exposure must fall by roughly 4-fold. The
# bound is deterministic (no cohort), but stated loosely because AUC over a
# finite 24 h window is not exactly inversely proportional to CL.
stopifnot(auc_off / auc_on > 2)
```

# Virtual cohort

Original observed data are not publicly available. The cohorts below are virtual
populations drawn from the model's own (log-normal approximated) parameter
distribution, stratified by the residual-diuresis values the paper itself
simulates: 0, 100 and 300 mL/24h.

```{r cohort}
# set.seed() seeds R's RNG, NOT rxode2's simulation RNG, whose streams are
# partitioned per solver thread -- so this cohort is reproducible on this machine
# and different on a machine with a different thread count. Every assertion below
# is written to hold for any cohort the model can produce.
set.seed(20260917)

n_arm <- 200L # cap is 200 participants per arm

rd_levels <- c(0, 100, 300)
events <- dplyr::bind_rows(lapply(seq_along(rd_levels), function(i) {
  make_events(n_arm, dose_mg = 1000, rd = rd_levels[i], tmax = 48, dt = 0.25,
              id_offset = (i - 1L) * n_arm) |>
    dplyr::mutate(arm = paste0("RD ", rd_levels[i], " mL/24h"))
}))

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(mod, events, keep = c("arm"), covsInterpolation = "locf",
                       returnType = "data.frame")
stopifnot(all(sim$Cc[!is.na(sim$Cc)] >= 0))
```

```{r cohort-profiles, echo = FALSE}
sim |>
  filter(!is.na(Cc)) |>
  group_by(arm, time) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
            .groups = "drop") |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~arm) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Meropenem (mg/L)",
       title = "Single 1 g IV dose over 30 min, no SLED session",
       caption = "Median and 5th-95th percentiles, 200 virtual subjects per arm.")
```

# Replicating Figure 5

Figure 5 of Braune 2018 shows the meropenem profile for **2 g 8-hourly in a
patient with a residual diuresis of 300 mL/d**, plotted as a percentile band over
roughly 56 hours. The published panel is a low-resolution raster without
readable axis annotations, so it is reproduced here qualitatively; the one
quantitative feature that can be read off it is the peak, which sits just below
the 250 mg/L gridline.

```{r figure-5}
# Replicates Figure 5 of Braune 2018. The paper's simulation places a 5-hour SLED
# session starting 17 h after the first dose (Methods, "Probability of target
# attainment"); the repeating sawtooth in Fig. 5 is q8h dosing.
ev_f5 <- make_events(n_arm, dose_mg = 2000, ii = 8, n_doses = 7L, rd = 300,
                     sled_start = 17, sled_dur = 5, tmax = 56, dt = 0.1)
sim_f5 <- rxode2::rxSolve(mod, ev_f5, covsInterpolation = "locf",
                          returnType = "data.frame")

sim_f5 |>
  filter(!is.na(Cc)) |>
  group_by(time) |>
  summarise(Q10 = quantile(Cc, 0.10), Q25 = quantile(Cc, 0.25), Q50 = median(Cc),
            Q75 = quantile(Cc, 0.75), Q90 = quantile(Cc, 0.90), .groups = "drop") |>
  tidyr::pivot_longer(-time, names_to = "pct", values_to = "Cc") |>
  ggplot(aes(time, Cc, group = pct)) +
  geom_line(alpha = 0.8) +
  annotate("rect", xmin = 17, xmax = 22, ymin = -Inf, ymax = Inf,
           alpha = 0.12, fill = "steelblue") +
  labs(x = "Time (h)", y = "Meropenem (mg/L)",
       title = "Figure 5 - 2 g 8-hourly, residual diuresis 300 mL/24h",
       caption = paste("Replicates Figure 5 of Braune 2018. Percentiles 10/25/50/75/90.",
                       "Shaded band = the 5-hour SLED session at 17-22 h."))
```

The peak amplitude is worth checking explicitly, because it is the one feature of
the published panel that carries information and because it is sensitive to a
mis-transcribed dose, volume or unit. Note that the *typical-value* peak is not
the top of the published band: distribution out of the central compartment is
very fast here (alpha half-life about 3 minutes, `Vp` about 5.7-fold `Vc`), so a
30-minute infusion never approaches `Dose / Vc`, and the upper percentile lines
in Figure 5 are driven by subjects who drew a small `Vc` (CV 87.9%).

```{r figure-5-amplitude}
# Closed form for a constant-rate infusion into a two-compartment system,
# evaluated at the end of the infusion (the peak of the first dose).
infusion_peak <- function(dose, tinf, vc, k12, k21, kel) {
  s <- kel + k12 + k21
  disc <- sqrt(s^2 - 4 * kel * k21)
  alpha <- (s + disc) / 2
  beta <- (s - disc) / 2
  r0 <- dose / tinf
  (r0 / vc) * (((alpha - k21) / (alpha - beta)) / alpha * (1 - exp(-alpha * tinf)) +
                 ((k21 - beta) / (alpha - beta)) / beta * (1 - exp(-beta * tinf)))
}

ev_peak <- make_events(1, dose_mg = 2000, rd = 300, tmax = 8, dt = 0.02)
sim_peak <- solve_typical(ev_peak)
peak_sim <- max(sim_peak$Cc, na.rm = TRUE)
peak_cf <- infusion_peak(2000, 0.5, p$vc, p$kcp, p$kpc, published_cl(300, 0) / p$vc)

cat(sprintf("first-dose typical peak: simulated %.2f mg/L, closed form %.2f mg/L (%.3f%%)\n",
            peak_sim, peak_cf, 100 * (peak_sim / peak_cf - 1)))
cat(sprintf("Dose/Vc (instantaneous-bolus ceiling, not reached): %.1f mg/L\n", 2000 / p$vc))

# Deterministic identity against an exact algebraic solution.
stopifnot(abs(peak_sim / peak_cf - 1) < 0.01)

# Descriptive: where the percentile bands of the Figure 5 cohort peak. The
# published panel's uppermost line rises to near the top of its axis; the
# corresponding simulated band is reported here rather than gated, because the
# published axis annotations are not legible at the distributed resolution.
band_peaks <- sim_f5 |>
  filter(!is.na(Cc)) |>
  group_by(time) |>
  summarise(across(Cc, list(p10 = \(x) quantile(x, 0.10), p50 = median,
                            p90 = \(x) quantile(x, 0.90))), .groups = "drop") |>
  summarise(across(-time, max))
print(round(as.data.frame(band_peaks), 1))
```

## Observed trough concentrations (context, not a gate)

Table 1 of Braune 2018 reports a median observed trough of 28.9 mg/L
[IQR 21.6-36.9, range 10.2-95.8]. That value pools the 0.5 g, 1 g and 2 g dose
levels with no per-patient dose breakdown, so it cannot be reconstructed from the
model and is **not** gated. It is shown here only to confirm the model predicts
troughs of the right order for the doses the cohort actually received.

```{r trough-context}
trough_ctx <- dplyr::bind_rows(lapply(c(500, 1000, 2000), function(d) {
  ev <- make_events(n_arm, dose_mg = d, ii = 8, n_doses = 4L, rd = 0,
                    sled_start = 17, sled_dur = 5, tmax = 32, dt = 0.2,
                    id_offset = d)
  s <- rxode2::rxSolve(mod, ev, covsInterpolation = "locf", returnType = "data.frame")
  tr <- s$Cc[!is.na(s$Cc) & s$time == 32]
  dplyr::tibble(dose = paste(d, "mg q8h"), Median = median(tr),
                Q1 = quantile(tr, 0.25), Q3 = quantile(tr, 0.75))
})) |>
  mutate(across(c(Median, Q1, Q3), \(x) round(x, 1)))

trough_ctx |>
  dplyr::rename("Regimen" = dose, "Simulated trough median (mg/L)" = Median,
                "Q1" = Q1, "Q3" = Q3) |>
  knitr::kable(caption = paste(
    "Simulated 32 h trough at residual diuresis 0 mL/24h, by dose level.",
    "Braune 2018 Table 1 reports an observed median trough of 28.9 mg/L",
    "[IQR 21.6-36.9] pooled across all three dose levels."
  ))
```

# PKNCA validation

## Deterministic NCA against the published parameters

The paper reports no NCA table, so the reference column here is built from the
published parameters themselves: clearance from the printed covariate equation
(Table 2 values `CLD = 2.6` and `CLN = 1.5`) and terminal half-life from the
closed-form eigenvalue of the Table 2 two-compartment system. Both sides use the
same drawn parameters, so the difference is numerical only and is gated tightly.

```{r pknca-typical}
nca_arms <- lapply(seq_along(rd_levels), function(i) {
  make_events(1, dose_mg = 1000, rd = rd_levels[i], tmax = 96, dt = 0.05,
              id_offset = i) |>
    mutate(arm = paste0("RD ", rd_levels[i], " mL/24h"))
})
sim_typ_nca <- dplyr::bind_rows(lapply(nca_arms, function(ev) {
  solve_typical(ev) |> mutate(arm = ev$arm[1], id = ev$id[1])
}))

conc_typ <- sim_typ_nca |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, arm)
conc_typ <- dplyr::bind_rows(
  conc_typ,
  conc_typ |> distinct(id, arm) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, arm, time, .keep_all = TRUE) |>
  arrange(id, arm, time)
stopifnot(nrow(conc_typ) > 0, all(conc_typ$Cc >= 0))

dose_typ <- dplyr::bind_rows(nca_arms) |>
  filter(evid == 1) |>
  transmute(id, time, amt, arm, route = "intravascular", duration = 0.5)

conc_obj <- PKNCA::PKNCAconc(conc_typ, Cc ~ time | arm + id)
dose_obj <- PKNCA::PKNCAdose(dose_typ, amt ~ time | arm + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE,
  cl.obs = TRUE, half.life = TRUE
)
nca_typ <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

# Closed-form terminal (beta) half-life from the Table 2 parameters.
beta_half_life <- function(rd) {
  kel <- published_cl(rd, 0) / p$vc
  s <- kel + p$kcp + p$kpc
  beta <- (s - sqrt(s^2 - 4 * kel * p$kpc)) / 2
  log(2) / beta
}

published_nca <- dplyr::tibble(
  arm = paste0("RD ", rd_levels, " mL/24h"),
  cl.obs = published_cl(rd_levels, 0),
  half.life = vapply(rd_levels, beta_half_life, numeric(1))
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_typ,
  reference = published_nca,
  by = "arm",
  params = c("cl.obs", "half.life"),
  units = c(cl.obs = "L/h", half.life = "h"),
  tolerance_pct = 5
)
knitr::kable(cmp, caption = paste(
  "Typical-value NCA versus the published clearance equation and the",
  "closed-form terminal half-life. * marks a difference above 5%."
))
```

```{r pknca-typical-gate}
res_typ <- as.data.frame(nca_typ)
pick <- function(par, a) {
  v <- res_typ$PPORRES[res_typ$PPTESTCD == par & res_typ$arm == a]
  if (length(v) != 1L) stop("no unique NCA row for ", par, " in arm ", a)
  v
}
gate <- dplyr::tibble(
  arm = published_nca$arm,
  cl_pct = 100 * (vapply(published_nca$arm, \(a) pick("cl.obs", a), numeric(1)) /
                    published_nca$cl.obs - 1),
  hl_pct = 100 * (vapply(published_nca$arm, \(a) pick("half.life", a), numeric(1)) /
                    published_nca$half.life - 1)
)
print(as.data.frame(gate), digits = 4)

# Deterministic: both sides use the same parameters, so only trapezoidal and
# lambda-z regression error separate them. Realised |cl.obs| <= 0.017% and
# |half.life| <= 0.064%; 0.5% is ~8-fold headroom over that and still goes red
# on any real transcription or structural regression.
stopifnot(nrow(gate) == 3L, !anyNA(gate$cl_pct), !anyNA(gate$hl_pct))
stopifnot(max(abs(gate$cl_pct)) < 0.5, max(abs(gate$hl_pct)) < 0.5)
```

## Population NCA summary

```{r pknca-cohort}
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, arm)
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> distinct(id, arm) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, arm, time, .keep_all = TRUE) |>
  arrange(id, arm, time)
stopifnot(nrow(sim_nca) > 0)

dose_cohort <- events |>
  filter(evid == 1) |>
  transmute(id, time, amt, arm, route = "intravascular", duration = 0.5)

nca_cohort <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id),
  PKNCA::PKNCAdose(dose_cohort, amt ~ time | arm + id),
  intervals = data.frame(start = 0, end = 48, cmax = TRUE, tmax = TRUE,
                         auclast = TRUE)
))

as.data.frame(nca_cohort) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "auclast")) |>
  group_by(arm, PPTESTCD) |>
  summarise(Median = median(PPORRES), Q1 = quantile(PPORRES, 0.25),
            Q3 = quantile(PPORRES, 0.75), .groups = "drop") |>
  mutate(across(c(Median, Q1, Q3), \(x) signif(x, 3))) |>
  dplyr::rename("Arm" = arm, "NCA parameter" = PPTESTCD,
                "Q1" = Q1, "Q3" = Q3) |>
  knitr::kable(caption = paste(
    "Population NCA after a single 1 g IV dose (no SLED session),",
    "200 virtual subjects per arm. Cmax mg/L, tmax h, AUC(0-48) mg*h/L."
  ))
```

# Probability of target attainment

The paper's headline output is the PTA for `40% fT>MIC` and `100% fT>MIC` over
the first 24 h against *P. aeruginosa*, computed by Monte Carlo from the same
population distribution. The simulation design is specified in Methods
("Probability of target attainment"): a 5-hour SLED session beginning 17 h after
the first dose, 30-minute infusions, and free concentrations taken as 98% of
total (meropenem is 2% protein bound).

```{r pta}
mic <- 2 # mg/L, the EUCAST susceptibility breakpoint the paper uses
fu <- 0.98 # 2% plasma protein binding (Braune 2018 Methods)

regimens <- dplyr::tribble(
  ~regimen, ~dose_mg, ~ii, ~n_doses,
  "0.5 g q8h", 500, 8, 3L,
  "1 g q12h", 1000, 12, 2L,
  "2 g q8h", 2000, 8, 3L
)

pta_grid <- tidyr::expand_grid(regimens, rd = rd_levels) |>
  mutate(arm_id = dplyr::row_number())

ev_pta <- dplyr::bind_rows(lapply(seq_len(nrow(pta_grid)), function(i) {
  g <- pta_grid[i, ]
  make_events(n_arm, dose_mg = g$dose_mg, ii = g$ii, n_doses = g$n_doses,
              rd = g$rd, sled_start = 17, sled_dur = 5, tmax = 24, dt = 0.1,
              id_offset = (g$arm_id - 1L) * n_arm) |>
    mutate(regimen = g$regimen, rd_lbl = g$rd)
}))
stopifnot(!anyDuplicated(unique(ev_pta[, c("id", "time", "evid")])))

sim_pta <- rxode2::rxSolve(mod, ev_pta, keep = c("regimen", "rd_lbl"),
                           covsInterpolation = "locf", returnType = "data.frame")

# fT>MIC over 0-24 h: the time-weighted fraction of the window on which the free
# concentration exceeds the MIC (a left-Riemann sum over the observation grid).
ft_mic <- sim_pta |>
  filter(!is.na(Cc)) |>
  group_by(regimen, rd_lbl, id) |>
  arrange(time, .by_group = TRUE) |>
  summarise(
    ft = {
      w <- diff(c(time, 24))
      above <- (fu * Cc) > mic
      sum(w * above) / 24
    },
    .groups = "drop"
  )

# The window starts at the first dose, so the opening instants of the first
# 30-minute infusion are necessarily below the MIC and an ft of exactly 1 is
# unattainable by construction. "100% fT>MIC" is therefore scored at 0.99.
pta <- ft_mic |>
  group_by(regimen, rd_lbl) |>
  summarise(`PTA 40% fT>MIC (%)` = 100 * mean(ft >= 0.40),
            `PTA 100% fT>MIC (%)` = 100 * mean(ft >= 0.99),
            .groups = "drop") |>
  mutate(across(where(is.numeric) & !dplyr::all_of("rd_lbl"), \(x) round(x, 1)))

pta |>
  dplyr::rename("Regimen" = regimen, "Residual diuresis (mL/24h)" = rd_lbl) |>
  knitr::kable(caption = paste(
    "Simulated probability of target attainment at MIC = 2 mg/L over the first",
    "24 h, with a 5-hour SLED session at 17-22 h. 200 virtual subjects per arm."
  ))
```

```{r pta-claims}
pta_at <- function(reg, rd, col) {
  v <- pta[[col]][pta$regimen == reg & pta$rd_lbl == rd]
  if (length(v) != 1L) stop("no unique PTA row for ", reg, " at RD ", rd)
  v
}

claims <- dplyr::tribble(
  ~Claim, ~Published, ~Simulated,
  "0.5 g q8h, RD 0, 40% fT>MIC", ">95",
  pta_at("0.5 g q8h", 0, "PTA 40% fT>MIC (%)"),
  "0.5 g q8h, RD 100, 40% fT>MIC", ">95",
  pta_at("0.5 g q8h", 100, "PTA 40% fT>MIC (%)"),
  "0.5 g q8h, RD 300, 40% fT>MIC", ">95",
  pta_at("0.5 g q8h", 300, "PTA 40% fT>MIC (%)"),
  "1 g q12h, RD 0, 100% fT>MIC", ">95",
  pta_at("1 g q12h", 0, "PTA 100% fT>MIC (%)"),
  "2 g q8h, RD 300, 100% fT>MIC", "93",
  pta_at("2 g q8h", 300, "PTA 100% fT>MIC (%)")
)

claims |>
  dplyr::rename("Braune 2018 claim (Results / Discussion)" = Claim,
                "Paper PTA (%)" = Published, "This model (%)" = Simulated) |>
  knitr::kable(caption = paste(
    "Braune 2018 prose PTA claims against the packaged model. The",
    "reconstruction runs systematically BELOW the paper throughout, by a few",
    "points at the 40% target and by tens of points at the 100% target; see",
    "the narrative and 'Assumptions and deviations' below."
  ))
```

The reconstruction is systematically low, and the size of the gap tracks how far
into the distribution's tails the target reaches. At the traditional
`40% fT>MIC` target the gap is a few points (85-94% here against the paper's
">95%"). At the aggressive `100% fT>MIC` target it is tens of points, and it is
worst where the clearance distribution matters most.

That is the expected signature of approximating a 19-subject non-parametric
support-point distribution by an independent log-normal with the same CV. A
log-normal at CV 134.7% (`CLN`) or 87.9% (`Vc`) puts far more mass in the
extreme tails than a discrete 19-point distribution can, and a target that
requires the concentration to stay above the MIC for the *entire* interval is
decided precisely by those tails. The gates below therefore pin the quantities
that are robust to this approximation -- a floor on the traditional target, and
the dose-separation the paper's dosing recommendations rest on -- and the
`100% fT>MIC` rows are recorded as a known deviation rather than gated.

```{r pta-gates}
stopifnot(nrow(pta) == 9L, !anyNA(pta$`PTA 40% fT>MIC (%)`),
          !anyNA(pta$`PTA 100% fT>MIC (%)`))

# 1. The traditional target is attained by most subjects on the lowest studied
#    regimen at every residual diuresis. Realised 93.5 / 90.5 / 85.0; the floor
#    of 70 leaves ~15 points of headroom for cohort noise while still going red
#    on a mis-transcribed dose, clearance or volume (which move these by tens of
#    points, not by a few).
pta40_low <- vapply(rd_levels, \(r) pta_at("0.5 g q8h", r, "PTA 40% fT>MIC (%)"),
                    numeric(1))
stopifnot(all(pta40_low > 70))

# 2. The dose separation the paper's recommendations rest on: at the aggressive
#    target and the highest residual diuresis, 2 g q8h must beat 0.5 g q8h by a
#    wide margin. Realised 60 points; gated at 20 as a magnitude, not as a bare
#    ordering of two noisy statistics.
stopifnot(pta_at("2 g q8h", 300, "PTA 100% fT>MIC (%)") -
            pta_at("0.5 g q8h", 300, "PTA 100% fT>MIC (%)") > 20)

# 3. Residual diuresis must cost a meaningful amount of target attainment at the
#    aggressive target on the lowest regimen. Realised 35.0 - 13.5 = 21.5 points.
stopifnot(pta_at("0.5 g q8h", 0, "PTA 100% fT>MIC (%)") -
            pta_at("0.5 g q8h", 300, "PTA 100% fT>MIC (%)") > 5)
```

# Assumptions and deviations

- **Non-parametric distribution approximated as log-normal.** Braune 2018 fits
  with Pmetrics NPAG, which estimates a discrete non-parametric distribution
  rather than a parametric OMEGA. Table 2 summarises it by mean, SD, CV% and
  median. The packaged model carries the **Mean** column as the typical value
  (the column the Discussion quotes) and converts the CV% into a log-normal
  random effect via `omega^2 = log(CV^2 + 1)`, uncorrelated because Table 2
  reports no correlations. This is an approximation in both directions: the
  log-normal has a heavier right tail than a 19-subject support-point set can
  have, and a log-normal whose median is `exp(mu)` has arithmetic mean
  `exp(mu + omega^2/2)`, which exceeds the published mean. The effect is largest
  for `CLN` (CV 134.7%). Consequently the PTA table above is a *reconstruction*
  of the paper's Monte Carlo analysis, not a reproduction of it. **It runs
  systematically below the published values**: at the traditional
  `40% fT>MIC` target the packaged model gives 93.5 / 90.5 / 85.0% for 0.5 g q8h
  at residual diuresis 0 / 100 / 300 mL/24h, against the paper's ">95%" at all
  three; at the aggressive `100% fT>MIC` target it gives 62% for 1 g q12h at
  RD 0 (paper: ">95%") and 73.5% for 2 g q8h at RD 300 (paper: 93%). The
  `100% fT>MIC` rows are recorded as known deviations rather than gated, because
  that target is decided by the extreme tails of the clearance and volume
  distributions, which is exactly where a parametric approximation and a
  19-subject non-parametric original differ most. **Do not use this model to
  regenerate the paper's dosing recommendations**; use it as a structural PK
  model and take the dosing conclusions from the paper. The same caveat applies
  to the sibling Pmetrics extractions `Tsai_2023_ceftriaxone` and
  `Duke_2024_cefazolin`.
- **"100% fT>MIC" is scored at 0.99, not 1.0.** The paper's PTA window opens at
  the first dose, so the opening instants of the first 30-minute infusion are
  necessarily below the MIC and an exact `fT>MIC` of 1 is unattainable by
  construction. The threshold is set at 0.99 of the 24-hour window.
- **NPAG typical values are marginal means, not a typical subject.** Feeding the
  Table 2 means into the structural model gives a valid amplitude comparator
  (Cmax, AUC, clearance) but not necessarily the shape of any real patient's
  profile. Amplitude quantities are gated above; profile-shape quantities are
  reported without gates.
- **Residual error magnitude is unpublished.** Methods state that additive
  (lambda) and exponential (gamma) error models were tested and Results that an
  additive model was used, but no value is given. The publisher's supplementary
  file set for PMC5791175 was retrieved and contains only the five figure images
  plus the equation image -- no Pmetrics model file and no error table. `addSd`
  is therefore `fixed(0)`. It was deliberately **not** substituted from the
  HPLC-UV assay precision reported in Methods (CV 9.6% / 3.9% / 2.2% at
  10 / 20 / 80 mg/L), which is an assay CV and not a fitted residual SD.
- **The SLED arm is additive.** The printed equation is
  `TVCL = CL_SLED * (SLED) + CL_NS`, so total clearance during a session at
  `RD = 0` is `7.9 + 2.6 = 10.5` L/h. One Discussion sentence reads loosely
  against this ("Compared to healthy volunteers (11-14 L/h), meropenem CL is
  significantly lower during SLED (7.9 L/h)"), which would suggest `CLSLED` is
  the *total* intradialytic clearance, i.e. a replacement rather than an additive
  rule. The printed equation governs: it is unambiguous in the publisher's
  typeset image, it is stated in the Results rather than the Discussion, and the
  paper's own decomposition sentence ("CL without SLED (CLNS) is composed of CLN
  (1.5 L/h) and CLD (2.6 L/h)") defines `CLNS` as the interdialytic total, which
  the equation then adds to `CLSLED`. Note that the sibling Pmetrics models from
  a different group (`Tsai_2023_ceftriaxone`, `Duke_2024_cefazolin`) use the
  opposite, *replacement* rule -- the two conventions coexist in this literature
  and the source's own conditional has to be read each time.
- **Peripheral volume and inter-compartmental clearance are derived, not
  published.** Braune 2018 parameterises distribution by `Kcp` / `Kpc` only. The
  model file computes `vp <- vc * k12 / k21` and `q <- k12 * vc` (typical 46.4 L
  and 83.4 L/h). These are algebraic identities, not new information -- but they
  are load-bearing: without them rxode2 silently collapses the model to one
  compartment. See "Structural verification" section 2.
- **`RRT_CRRT_ACTIVE`, not `RRT_HEMODIAL_ACTIVE`.** SLED is a prolonged
  *intermittent* modality, but the register assigns it to the continuous /
  extended `RRT_CRRT_*` family, whose entry names SLED explicitly. The gate is
  time-varying within subject, so every `rxSolve()` call in this vignette passes
  `covsInterpolation = "locf"`; the rxode2 default interpolates covariates
  linearly, which would ramp a binary on/off indicator across the transition
  instead of switching it.
- **SLED session timing in the simulations** follows the paper's own Monte Carlo
  design (5 hours, starting 17 h after the first dose), not the observed sessions
  (median 315 min, range 80-470 min), so that the PTA reconstruction is
  comparable to the published one.
- **No weight scaling.** Body weight was screened and not retained, so all
  clearances and volumes are absolute rather than per-kilogram, despite the
  cohort's wide weight range (70-183 kg). Documented in `covariatesDataExcluded`.
- **Observed trough concentrations are not used as a gate.** Table 1 reports a
  median observed trough of 28.9 mg/L [IQR 21.6-36.9], but it pools the 0.5 g,
  1 g and 2 g dose levels with no per-patient dose breakdown and mixes samples
  taken before and around SLED sessions. It is an empirical summary over an
  unreported dose mix, so it cannot be reconstructed from the model and is
  recorded here as context only.
