---
title: "Pamrevlumab (FG-3019) target-mediated disposition in rats (Brenner 2016)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pamrevlumab (FG-3019) target-mediated disposition in rats (Brenner 2016)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Brenner_2016_pamrevlumab_rat"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1007/s11095-016-1918-0> (open access; PMCID PMC4942499)
* Supplement (model equations): Electronic Supplementary Material to the above, "Kinetic Model development", Eqs. (1)-(20).

FG-3019 is the development code for the human anti-CTGF IgG1 monoclonal
antibody later assigned the INN **pamrevlumab**; the paper uses "FG-3019"
throughout, and this vignette does too when quoting the source.

# Population

The model was fit to three data sets from male Sprague-Dawley rats (average
336 g at dosing; the paper expresses model parameters for a 0.3 kg animal),
all fit **simultaneously** by maximum likelihood in ADAPT 5:

1. **FG-3019 single-dose PK** (Fig. 1) -- IV bolus at 0.03, 0.3, 3, 10, 30 and
   100 mg/kg (n = 3, 3, 9, 6, 6, 6), sampled to 504 h.
2. **Recombinant human CTGF and CTGF N-fragment PK** (Fig. 3) -- IV bolus at
   20 and 40 nmol/kg of each form, 3 rats per group. The human-specific assays
   do not detect endogenous rat CTGF, so the paper set the baselines to zero for
   these data.
3. **Endogenous CTGF-N response to FG-3019** (Fig. 2) -- the N+W-CTGF assay
   measured in the 10, 30 and 100 mg/kg FG-3019 groups.

Mean (not individual) data were fit, so the CV% values in Table I are
**estimation precision, not between-animal variability**. The model therefore
carries no IIV and no residual error and is a typical-value mechanistic
simulator.

The co-administration experiments (Figs. 4-5) were deliberately excluded from
the fit by the authors: the model treats FG-3019 as a monovalent 75 kDa species,
whereas co-dosed rhCTGF forms 2:1 complexes with the bivalent 150 kDa antibody,
and no rat RAP kinetics were available.

```{r population}
str(ui$population)
```

# Source trace

Every `ini()` entry carries an in-file comment pointing at its source location.
Collected here for review.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lvc` (V) | 0.01512 L | Table I, CV 6.3% |
| `clw` (CLW) | 0.09243 L/h | Table I, CV 3.7% |
| `vwt` (VWT) | 0.1160 L | Table I, CV 13.0% |
| `clwt` (CLWT) | 1.380 L/h | Table I, CV 9.5% |
| `cldw` (CLdW) | 10 L/h | Table I, FIXED |
| `vnt` (VNT) | 0.04458 L | Table I, CV 11.5% |
| `cldn` (CLdN) | 0.07998 L/h | Table I, CV 9.4% |
| `kdw` (KDW) | 23.97 nM | Table I, CV 18.0% |
| `kdn` (KDN) | 51.21 nM | Table I, CV 15.0% |
| `koffw`, `koffn` | 100 1/h | Table I, FIXED |
| `bl_ctgf` (CW0) | 0.02591 nM | Table I, CV 15.6% |
| `bl_ctgfn` (CN0) | 0.6572 nM | Table I, CV 10.6% |
| `clab` (CLAb) | 0.0001321 L/h | Table I, CV 6.4% |
| `vabt` (VAbT) | 0.01363 L | Table I, CV 12.5% |
| `cldab` (CLdAb) | 0.0005692 L/h | Table I, CV 27.7% |
| `cldabw` (CLdAbW) | 10 L/h | Table I, FIXED |
| `cln = clw` | derived | Table I footnote a |
| `clabw = clabn = clab` | derived | Table I footnote b |
| `cldabn = cldab` | derived | Table I footnote c |
| `clabwt = clwt` | derived | Table I footnote d |
| `vabwt = vwt`, `vabnt = vabt` | derived | Methods, "Pharmacokinetic Modeling of Target-Mediated Elimination" |
| `konw`, `konn` | derived | Supplement Eq. (15), `kon = koff / KD` |
| `kw`, `kn` | derived | Supplement Eqs. (11)-(12); Table I footnote e ("secondary parameter") |
| Initial conditions | derived | Supplement Eqs. (13), (14), (16) |
| `d/dt(target_ctgf)`, `d/dt(target_ctgf_peripheral1)` | n/a | Supplement Eqs. (1)-(2) |
| `d/dt(target_ctgfn)`, `d/dt(target_ctgfn_peripheral1)` | n/a | Supplement Eqs. (3)-(4) |
| `d/dt(central)`, `d/dt(peripheral1)` | n/a | Supplement Eqs. (5)-(6) |
| `d/dt(complex_ctgf)`, `d/dt(complex_ctgf_peripheral1)` | n/a | Supplement Eqs. (7)-(8) |
| `d/dt(complex_ctgfn)`, `d/dt(complex_ctgfn_peripheral1)` | n/a | Supplement Eqs. (9)-(10) |
| `d/dt(elim_target)`, `d/dt(elim_nontarget)` | n/a | Supplement Eqs. (17)-(20) |

## Table I footnote e is reproducible, and that checks the transcription

`kW` and `kN` are tabulated as *secondary* parameters -- the paper computed them
from the baselines via Supplement Eqs. (11)-(12) rather than estimating them
directly. The model file derives them the same way, so substituting Table I's
values must return Table I's own numbers. It does, to four significant figures,
which independently confirms that the steady-state algebra (and hence the
transcription of Eqs. 1-4) is right.

```{r footnote-e}
p <- as.list(setNames(ui$theta, names(ui$theta)))
vc <- exp(p$lvc)
kw_paper <- 0.03382; kn_paper <- 0.06075
kw_model <- p$bl_ctgf * (p$clw + p$cldw * p$clwt / (p$cldw + p$clwt))  # Eq. (11)
kn_model <- p$clw * p$bl_ctgfn                                         # Eq. (12), CLN = CLW
c(kw_model = kw_model, kw_paper = kw_paper,
  kn_model = kn_model, kn_paper = kn_paper)
stopifnot(
  abs(kw_model - kw_paper) / kw_paper < 1e-3,
  abs(kn_model - kn_paper) / kn_paper < 1e-3
)
```

# Unit conventions and dosing

* Time: hours. States: amounts in nmol. Concentrations: nM. Volumes: L.
* FG-3019 is counted in **binding sites**: MW 75 kDa, i.e. half of the 150 kDa
  bivalent IgG, so 1 mg = 1000/75 = 13.33 nmol of binding sites.
* Intact CTGF MW 38 kDa; CTGF N-fragment MW 19 kDa.
* All model parameters are expressed for a **0.3 kg** rat (the paper's own
  normalisation: V = 0.01512 L = 50.4 mL/kg at 0.3 kg).

```{r units}
mod <- readModelDb("Brenner_2016_pamrevlumab_rat")
wt_kg <- 0.3                  # Brenner 2016 Results: "for a 0.3 kg animal"
nmol_per_mg_ab <- 1000 / 75   # FG-3019 binding sites, MW 75 kDa
ug_per_mL_per_nM <- 75 / 1000 # 1 nM binding sites = 0.075 ug/mL

# The paper's own consistency check: 3 mg/kg should give a Cmax near 1 uM.
c0_3mgkg <- 3 * wt_kg * nmol_per_mg_ab / exp(p$lvc)
c0_3mgkg  # nM; paper: "approximately 1 uM"
stopifnot(c0_3mgkg > 600, c0_3mgkg < 1400)

solve_fg <- function(dose_mgkg, times) {
  ev <- rxode2::et(amt = dose_mgkg * wt_kg * nmol_per_mg_ab,
                   cmt = "central", time = 0) |>
    rxode2::et(times)
  s <- as.data.frame(rxode2::rxSolve(mod, ev))
  if (is.null(s$id)) s$id <- 1L
  s$dose_mgkg <- dose_mgkg
  s
}
```

# Validation 1 -- the endogenous system holds at baseline

With no antibody dose, the CTGF and CTGF-N states must sit exactly at the
baselines of Supplement Eqs. (13), (14) and (16) and never drift. This is the
sharpest available test of the production rates, the tissue initial conditions
and the sign of every exchange term simultaneously: any error in Eqs. (1)-(4)
shows up as drift.

```{r steady-state}
ss <- as.data.frame(rxode2::rxSolve(mod, rxode2::et(seq(0, 500, by = 1))))
ss_summary <- tibble(
  quantity = c("Cctgf_w (intact CTGF)", "Cctgf_nw (N+W assay)", "Cc (antibody)"),
  expected = c(p$bl_ctgf, p$bl_ctgf + p$bl_ctgfn, 0),
  min = c(min(ss$Cctgf_w), min(ss$Cctgf_nw), min(ss$Cc)),
  max = c(max(ss$Cctgf_w), max(ss$Cctgf_nw), max(ss$Cc))
)
knitr::kable(ss_summary, digits = 8,
             caption = "Baseline hold over 500 h with no dose (nM).")

# Deterministic model, no IIV and no RNG anywhere: drift is pure solver error,
# so this bound is tight on purpose (realised ~1e-16).
stopifnot(
  diff(range(ss$Cctgf_w)) < 1e-9,
  diff(range(ss$Cctgf_nw)) < 1e-9,
  max(ss$Cc) < 1e-12,
  abs(max(ss$Cctgf_w) - p$bl_ctgf) < 1e-9,
  abs(max(ss$Cctgf_nw) - (p$bl_ctgf + p$bl_ctgfn)) < 1e-9
)
```

# Validation 2 -- mass balance of the eliminated antibody

Supplement Eqs. (17)-(20) split FG-3019 elimination into a target-mediated
route (tissue clearance of the antibody-CTGF complex) and a non-target-mediated
route. Integrated to depletion, the two must account for the **entire** dose.
This checks that no antibody mass leaks from, or is created by, the ten-state
system.

```{r mass-balance}
tgrid_long <- c(seq(0, 100, by = 0.5), seq(101, 3000, by = 5))
doses <- c(0.03, 0.3, 3, 10, 30, 100)

balance <- lapply(doses, function(d) {
  s <- solve_fg(d, tgrid_long)
  last <- s[nrow(s), ]
  tibble(
    dose_mgkg = d,
    dose_nmol = d * wt_kg * nmol_per_mg_ab,
    eliminated_nmol = last$elim_target + last$elim_nontarget,
    pct_target_mediated = last$pctTargetMediated
  )
}) |> bind_rows() |>
  mutate(pct_recovered = 100 * eliminated_nmol / dose_nmol)

knitr::kable(balance, digits = 3,
             caption = "Dose recovery and elimination-pathway split, integrated to 3000 h.")

stopifnot(all(abs(balance$pct_recovered - 100) < 0.1))
```

# Validation 3 -- the target-mediated pathway is structurally live

A model can reproduce a profile while silently ignoring the mechanism it was
built to express. Re-solving with the complex's tissue elimination driven to
zero (`clwt`, which also sets `clabwt` via Table I footnote d) must change the
antibody profile substantially at a low dose, where the paper says
target-mediated clearance dominates.

```{r structural-gate}
ev_low <- rxode2::et(amt = 0.03 * wt_kg * nmol_per_mg_ab, cmt = "central", time = 0) |>
  rxode2::et(seq(0, 400, by = 1))
s_on  <- as.data.frame(rxode2::rxSolve(mod, ev_low))
s_off <- as.data.frame(rxode2::rxSolve(mod, ev_low, params = c(clwt = 1e-9)))

rel_change <- max(abs(s_off$Cc - s_on$Cc) / pmax(s_on$Cc, 1e-12))
rel_change

# Deterministic; removing the dominant elimination route at 0.03 mg/kg changes
# the profile by orders of magnitude, so a 10-fold floor cannot be met by noise.
stopifnot(rel_change > 10)
```

# Replicate Figure 1 -- FG-3019 dose-ranging PK

```{r figure-1, fig.cap = "Replicates Figure 1 of Brenner 2016: FG-3019 binding-site concentration after IV bolus doses of 0.03-100 mg/kg."}
tgrid_fig <- c(seq(0, 24, by = 0.25), seq(25, 1200, by = 2))
fig1 <- lapply(doses, function(d) solve_fg(d, tgrid_fig)) |> bind_rows()

fig1 |>
  filter(Cc > 0.01) |>
  mutate(dose = factor(paste0(dose_mgkg, " mg/kg"),
                       levels = paste0(doses, " mg/kg"))) |>
  ggplot(aes(time, Cc, colour = dose)) +
  geom_line(linewidth = 0.7) +
  scale_y_log10(limits = c(0.01, 1e5)) +
  labs(x = "Time (h)", y = "FG-3019 binding sites (nM)", colour = NULL,
       caption = "Replicates Figure 1 of Brenner 2016.") +
  theme_bw()
```

The characteristic target-mediated shape of the published figure is reproduced:
a rapid distribution phase, a concentration-dependent plateau at high dose while
the target-mediated route is saturated, and a steeper terminal phase once it is
not. As Brenner 2016 notes, the true terminal phase lies well beyond the 504 h
sampling window -- which is why the published non-compartmental half-lives rise
with dose even though a TMDD model has a dose-independent terminal slope.

# Replicate Figure 3 -- recombinant human CTGF and CTGF-N PK

The human-specific assays do not detect endogenous rat CTGF, so the paper set
`W0 = N0 = 0` for these data. Setting both baselines to (effectively) zero also
zeroes the production rates, because `kw` and `kn` are derived from them via
Supplement Eqs. (11)-(12) -- exactly the scenario the paper fit.

```{r figure-3, fig.cap = "Replicates Figure 3 of Brenner 2016: rhCTGF-N (top) and rhCTGF (bottom) after IV bolus doses of 20 and 40 nmol/kg."}
zero_baselines <- c(bl_ctgf = 1e-12, bl_ctgfn = 1e-12)
tgrid_ctgf <- c(seq(0, 2, by = 0.002), seq(2.05, 24, by = 0.05))

solve_ctgf <- function(amt_nmol, cmt) {
  ev <- rxode2::et(amt = amt_nmol, cmt = cmt, time = 0) |> rxode2::et(tgrid_ctgf)
  s <- as.data.frame(rxode2::rxSolve(mod, ev, params = zero_baselines))
  if (is.null(s$id)) s$id <- 1L
  s
}

fig3 <- bind_rows(
  solve_ctgf(20 * wt_kg, "target_ctgfn") |> mutate(species = "rhCTGF-N", dose = "20 nmol/kg", conc = Cctgf_nw),
  solve_ctgf(40 * wt_kg, "target_ctgfn") |> mutate(species = "rhCTGF-N", dose = "40 nmol/kg", conc = Cctgf_nw),
  solve_ctgf(20 * wt_kg, "target_ctgf")  |> mutate(species = "rhCTGF",   dose = "20 nmol/kg", conc = Cctgf_w),
  solve_ctgf(40 * wt_kg, "target_ctgf")  |> mutate(species = "rhCTGF",   dose = "40 nmol/kg", conc = Cctgf_w)
)

fig3 |>
  filter(conc > 1e-4) |>
  ggplot(aes(time, conc, colour = dose)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~species, scales = "free") +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Concentration (nM)", colour = NULL,
       caption = "Replicates Figure 3 of Brenner 2016.") +
  theme_bw()
```

## Terminal half-lives of the two CTGF forms

Brenner 2016 reports a 43.5 min terminal half-life for rhCTGF-N and a 3.3 min
terminal half-life for intact rhCTGF -- the 13-fold difference that motivates
the whole model. Both are recovered from the packaged model by regressing the
log-linear terminal phase.

```{r ctgf-halflife}
tail_slope <- function(df, conc, lo, hi) {
  # Keep conc >= 1e-6 * Cmax: below that the ODE integrator has no relative accuracy left.
  d <- df[df$time >= lo & df$time <= hi & df[[conc]] >= 1e-6 * max(df[[conc]], na.rm = TRUE), ]
  unname(coef(lm(log(d[[conc]]) ~ d$time))[2])
}
sN <- solve_ctgf(20 * wt_kg, "target_ctgfn")
sW <- solve_ctgf(20 * wt_kg, "target_ctgf")

hl <- tibble(
  species = c("rhCTGF-N", "rhCTGF (intact)"),
  model_min = c(log(2) / -tail_slope(sN, "Cctgf_nw", 2, 6) * 60,
                log(2) / -tail_slope(sW, "Cctgf_w", 0.3, 1) * 60),
  paper_min = c(43.5, 3.3)
) |>
  mutate(pct_diff = 100 * (model_min - paper_min) / paper_min)

knitr::kable(hl, digits = 2,
             caption = "Terminal half-life of each CTGF form vs Brenner 2016 Results.")

# Deterministic. The model's true terminal slope vs the paper's NCA estimate on
# sampled mean data; 20% admits that difference, not cohort noise.
stopifnot(all(abs(hl$pct_diff) < 20))
# The 13-fold separation between the two forms is the paper's central claim.
stopifnot(hl$model_min[1] / hl$model_min[2] > 8)
```

# Replicate Figure 2 -- endogenous CTGF-N accumulates after FG-3019

Binding FG-3019 raises the apparent molecular weight of CTGF-N above the renal
filtration cut-off, so the complex is cleared ~700-fold more slowly than free
CTGF-N and the N+W-CTGF signal rises, peaks, and returns to baseline.

```{r figure-2, fig.cap = "Replicates Figure 2 of Brenner 2016: N+W-CTGF after FG-3019 at 10, 30 and 100 mg/kg."}
fig2 <- lapply(c(10, 30, 100), function(d) solve_fg(d, seq(0, 504, by = 1))) |>
  bind_rows() |>
  mutate(dose = factor(paste0(dose_mgkg, " mg/kg"),
                       levels = paste0(c(10, 30, 100), " mg/kg")))

ggplot(fig2, aes(time, Cctgf_nw, colour = dose)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time (h)", y = "N+W-CTGF (nM)", colour = NULL,
       caption = "Replicates Figure 2 of Brenner 2016.") +
  theme_bw()
```

Brenner 2016's Figure 2 plots this response on an **nM** axis (0-160 nM), and
the peaks of the published fitted curves read approximately 19, 38 and 86 nM at
10, 30 and 100 mg/kg. The packaged model reproduces them.

```{r figure-2-peaks}
fig2_peaks <- fig2 |>
  group_by(dose_mgkg) |>
  summarise(cmax_nM = max(Cctgf_nw), tmax_h = time[which.max(Cctgf_nw)], .groups = "drop") |>
  mutate(fig2_cmax_nM = c(19, 38, 86),
         pct_diff = 100 * (cmax_nM - fig2_cmax_nM) / fig2_cmax_nM)

knitr::kable(fig2_peaks, digits = 1,
             caption = "Peak N+W-CTGF vs the fitted curves read off Brenner 2016 Figure 2.")

# Reference values are read off a published figure, so ~15% covers the reading
# error; a mis-transcribed production rate or complex clearance moves these by
# a factor, not by 15%.
stopifnot(all(abs(fig2_peaks$pct_diff) < 15))
# Dose-ordered peak and dose-ordered delay, both stated in the Results text.
stopifnot(
  all(diff(fig2_peaks$cmax_nM) > 0),
  all(diff(fig2_peaks$tmax_h) > 0)
)
```

**Deviation (see Errata).** The Results *text* quotes "Cmax N+W-CTGF
concentrations of 32, 77 and 197 ng/ml" for these same three dose groups. Those
values cannot be reconciled with Figure 2's own nM axis under either molecular
weight in play (19 kDa for the N-fragment, 38 kDa for intact CTGF): the figure's
peaks correspond to roughly 360, 720 and 1630 ng/mL at 19 kDa. The model
reproduces the figure, which is the graphical record of both the data and the
fit, so the gate above is written against the figure and the text values are
recorded as an internal inconsistency in the source.

# Replicate Figure 7 -- dose-dependence of target-mediated elimination

This is the paper's most quantitative single claim and the sharpest available
check on the whole structure: at 100 mg/kg, **7.4%** of the dose is eliminated
by the target-mediated pathway, while at doses at or below 3 mg/kg that pathway
is the *major* route.

```{r figure-7, fig.cap = "Replicates Figure 7 of Brenner 2016: percent of the FG-3019 dose eliminated by the target-mediated pathway."}
ggplot(balance, aes(dose_mgkg, pct_target_mediated)) +
  geom_line(linewidth = 0.7) + geom_point() +
  geom_hline(yintercept = 50, linetype = "dashed", colour = "grey50") +
  scale_x_log10() +
  labs(x = "FG-3019 dose (mg/kg)", y = "Target-mediated elimination (%)",
       caption = "Replicates Figure 7 of Brenner 2016.") +
  theme_bw()
```

```{r figure-7-gate}
pct_100 <- balance$pct_target_mediated[balance$dose_mgkg == 100]
pct_le3 <- balance$pct_target_mediated[balance$dose_mgkg <= 3]
c(model_pct_at_100mgkg = pct_100, paper_pct_at_100mgkg = 7.4)

# Deterministic and directly quoted in the Results; 1 percentage point is tight
# but comfortable (realised difference ~0.1).
stopifnot(abs(pct_100 - 7.4) < 1)
# "For doses at or below 3 mg/kg, target-mediated clearance is the major pathway"
stopifnot(all(pct_le3 > 50))
# Monotone decreasing with dose, as Figure 7 shows.
stopifnot(all(diff(balance$pct_target_mediated) < 0))
```

# PKNCA validation against Table S1

Table S1 of the supplement reports non-compartmental parameters for each
FG-3019 dose group. To compare like with like, the model is sampled on a
schedule matching the study design (one pre-dose plus 12 post-dose points to
504 h) and analysed with PKNCA in the paper's units (ug/mL, days).

```{r pknca}
# Study-design sampling schedule (Brenner 2016 Methods: 12 post-dose points to 504 h).
sample_h <- c(0, 0.083, 0.5, 1, 6, 24, 48, 72, 120, 168, 240, 336, 504)

nca_sim <- lapply(doses, function(d) solve_fg(d, sample_h)) |>
  bind_rows() |>
  transmute(
    id        = as.integer(factor(dose_mgkg, levels = doses)),
    treatment = factor(paste0(dose_mgkg, " mg/kg"), levels = paste0(doses, " mg/kg")),
    time      = time / 24,                       # days
    Cc        = Cc * ug_per_mL_per_nM            # ug/mL
  )

# Only !is.na(); dropping time == 0 would remove PKNCA's AUC anchor.
nca_sim <- nca_sim |> filter(!is.na(Cc)) |>
  # Per subject, keep Cc >= 1e-6 * Cmax after the peak: below that the ODE integrator has no relative accuracy left.
  group_by(id) |>
  filter(time <= time[which.max(Cc)] | Cc >= 1e-6 * max(Cc)) |>
  ungroup()
stopifnot(nrow(nca_sim) > 0, any(nca_sim$time == 0))

dose_df <- tibble(
  id        = as.integer(factor(doses, levels = doses)),
  treatment = factor(paste0(doses, " mg/kg"), levels = paste0(doses, " mg/kg")),
  time      = 0,
  amt       = doses * wt_kg                      # mg administered
)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(nca_sim), Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | treatment + id,
                             route = "intravascular")

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE, cl.obs = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r nca-compare}
published <- tibble::tribble(
  ~treatment,     ~cmax,  ~aucinf.obs, ~half.life,
  "0.03 mg/kg",   0.453,  0.34,        1.36,
  "0.3 mg/kg",    5.75,   7.50,        1.63,
  "3 mg/kg",      63.3,   207,         1.70,
  "10 mg/kg",     292,    611,         2.62,
  "30 mg/kg",     828,    2490,        3.92,
  "100 mg/kg",    2665,   11300,       7.31
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by        = "treatment",
  units     = c(cmax = "ug/mL", aucinf.obs = "day*ug/mL", half.life = "day"),
  tolerance_pct = 20
)

knitr::kable(cmp, caption = "Simulated vs Brenner 2016 Table S1. * differs by >20%.")
```

The comparison is informative but is **not** a like-for-like test of the model,
for a reason the paper states itself: "sample collection from animals
administered high doses of FG-3019 was terminated too soon to establish the true
terminal elimination rates. This failure to collect samples at sufficiently late
time points explains the apparent dose-dependence of the experimentally
determined half-lives." Table S1's half-lives and extrapolated AUCs therefore
carry a dose-dependent truncation bias that the model, which has a
dose-independent true terminal slope, does not share. The rows are expected to
diverge at the high doses and are reported rather than gated.

What *is* gated is the non-linearity that the paper drew its conclusion from:
clearance must fall, and dose-normalised exposure must rise, monotonically with
dose.

```{r nca-trend}
nca_tab <- as.data.frame(nca_res) |>
  filter(PPTESTCD %in% c("cl.obs", "aucinf.obs")) |>
  select(treatment, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  mutate(
    dose_mgkg   = doses[match(treatment, paste0(doses, " mg/kg"))],
    cl_mL_day_kg = cl.obs * 1000 / wt_kg,
    auc_per_dose = aucinf.obs / dose_mgkg
  ) |>
  arrange(dose_mgkg)

nca_tab |>
  select(treatment, cl_mL_day_kg, auc_per_dose) |>
  dplyr::rename(
    "Dose group"                          = treatment,
    "CL (mL/day/kg)"                      = cl_mL_day_kg,
    "AUCinf/Dose (day*ug/mL per mg/kg)"   = auc_per_dose
  ) |>
  knitr::kable(digits = 1,
               caption = "Model non-linearity: CL falls and dose-normalised AUC rises with dose.")

# Deterministic solves, no cohort: assert the monotonicity directly.
stopifnot(
  all(diff(nca_tab$cl_mL_day_kg) < 0),
  all(diff(nca_tab$auc_per_dose) > 0)
)
# Brenner 2016 Results: CL falls from 90.6 to 8.9 mL/kg/day, a ~10-fold span.
stopifnot(
  nca_tab$cl_mL_day_kg[1] / nca_tab$cl_mL_day_kg[nrow(nca_tab)] > 4
)
```

# Assumptions and deviations

* **No IIV and no residual error.** Brenner 2016 fit *mean* data by maximum
  likelihood in ADAPT 5 and reports only estimation precision (CV%) in Table I.
  No between-animal variance and no residual-error model are published, so none
  is encoded; the model is a typical-value mechanistic simulator. The CV% values
  are preserved in the `ini()` source-trace comments.
* **Constrained parameters are derived, not duplicated.** Table I footnotes a-d
  record equality constraints (`CLN = CLW`; `CLAb = CLAbW = CLAbN`;
  `CLdAbN = CLdAb`; `CLAbWT = CLWT`) and the Methods add `VAbWT = VWT` and
  `VAbNT = VAbT`. These are computed in `model()` from the free parameters so
  the constraints cannot drift apart.
* **`CLAbW` typographical inconsistency in Table I.** Table I prints
  `CLAbW = 0.0001331 L/h` carrying footnote b, but footnote b itself states
  `CLAb = CLAbW = CLAbN`, and both `CLAb` and `CLAbN` are printed as
  `0.0001321 L/h`. The footnote is the constraint actually imposed during the
  fit and two of the three printed values agree with it, so `0.0001321` is used
  for all three and the printed `0.0001331` is treated as a typesetting error.
  The effect either way is below 1%.
* **Figure 2 vs the Results text.** The Results text quotes peak N+W-CTGF
  concentrations of "32, 77 and 197 ng/ml" at 10, 30 and 100 mg/kg. These are
  irreconcilable with Figure 2's own nM axis (peaks near 19, 38 and 86 nM, i.e.
  roughly 360, 720 and 1630 ng/mL at the N-fragment's 19 kDa). The packaged
  model reproduces the figure to within about 6%; the figure is used as the
  validation target and the text values are recorded here as an internal
  inconsistency in the source. No parameter was adjusted.
* **Table S1 NCA is not gated.** See the PKNCA section: the published
  non-compartmental parameters carry a dose-dependent terminal-phase truncation
  bias that the authors describe explicitly. The comparison table is rendered
  for transparency; the gated claim is the direction and magnitude of the
  non-linearity.
* **Experiments excluded by the authors.** The CTGF and RAP co-administration
  studies (Figs. 4-5) were not part of the fit -- the model treats FG-3019 as
  monovalent 75 kDa, whereas co-dosed rhCTGF forms 2:1 complexes with the
  bivalent antibody, and rat RAP kinetics were unavailable. They are therefore
  not reproduced here.
* **Fitting the human-CTGF experiments.** For Figure 3 the paper set
  `W0 = N0 = 0` because the human-specific assays do not detect endogenous rat
  CTGF. This vignette reproduces that by setting both baselines to 1e-12 nM,
  which also zeroes the derived production rates `kw` and `kn`.
* **Species.** All parameters are rat (Sprague-Dawley) values normalised to a
  0.3 kg animal. The paper notes non-linear FG-3019 PK was also seen in monkeys
  and humans, but reports no parameters for those species.
* Every parameter value comes from the paper's Table I or its Electronic
  Supplementary Material. No value was taken from any other source, digitised
  from a figure, or supplied by correspondence.
