---
title: "Liposomal irinotecan and SN-38 PK with exposure-safety (Brendel 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Liposomal irinotecan and SN-38 PK with exposure-safety (Brendel 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Source and models

Brendel K, Bekaii-Saab T, Boland PM, et al. Population pharmacokinetics of
liposomal irinotecan in patients with cancer and exposure-safety analyses in
patients with metastatic pancreatic cancer. *CPT Pharmacometrics Syst
Pharmacol.* 2021;10:1550-1563.
[doi:10.1002/psp4.12725](https://doi.org/10.1002/psp4.12725)

The paper contributes five models, all documented in this article:

| Model | What it is |
|---|---|
| `Brendel_2021_irinotecan_liposomal` | Joint popPK model of total irinotecan and SN-38 (Table 2, final NONMEM code in the Supplementary Material) |
| `Brendel_2021_irinotecan_liposomal_diarrhea_irinotecan` | Logistic model, grade >= 3 diarrhea vs log10 total-irinotecan Cavg,ss (Figure 4a) |
| `Brendel_2021_irinotecan_liposomal_diarrhea_sn38` | Logistic model, grade >= 3 diarrhea vs log10 SN-38 Cavg,ss (Figure 4b) |
| `Brendel_2021_irinotecan_liposomal_neutropenia_irinotecan` | Logistic model, grade >= 3 neutropenia AEs vs log10 total-irinotecan Cavg,ss (Figure 5a) |
| `Brendel_2021_irinotecan_liposomal_neutropenia_sn38` | Logistic model, grade >= 3 neutropenia AEs vs log10 SN-38 Cavg,ss (Figure 5b) |

An earlier analysis of 353 of the same patients, `Adiwijaya_2017_irinotecan_liposomal`,
fits the two analytes sequentially and with a different structure; Brendel
2021 replaced it with this joint parent-metabolite model.

```{r models}
mod_pk <- readModelDb("Brendel_2021_irinotecan_liposomal")
er_names <- c(
  diarrhea_irinotecan = "Brendel_2021_irinotecan_liposomal_diarrhea_irinotecan",
  diarrhea_sn38 = "Brendel_2021_irinotecan_liposomal_diarrhea_sn38",
  neutropenia_irinotecan = "Brendel_2021_irinotecan_liposomal_neutropenia_irinotecan",
  neutropenia_sn38 = "Brendel_2021_irinotecan_liposomal_neutropenia_sn38"
)
mod_er <- lapply(er_names, readModelDb)
ui_pk <- rxode2::rxode(mod_pk)
```

# Population

The popPK model was fit to 440 patients from seven studies (Table 1): the
phase III NAPOLI-1 trial (N = 260, second-line metastatic pancreatic ductal
adenocarcinoma, mPDAC), the phase I/II first-line mPDAC study NCT02551991
(N = 56, with 5-FU/LV and oxaliplatin), and five phase I/II studies in
various solid tumours or metastatic colorectal cancer (N = 124). Doses ranged
from 35 to 156 mg/m^2 irinotecan free base every 2 or 3 weeks, given as a
90-minute infusion. Supplementary Table S1: median age 62 years (28-87),
median BSA 1.71 m^2 (1.29-2.48), 48.6% female, 35.2% Asian, median bilirubin
0.41 mg/dL, median creatinine clearance 85 mL/min, 12.7% on oxaliplatin, 6.1%
UGT1A1*28 7/7 homozygous, and 18.6% treated with product from the previous
manufacturing site.

The exposure-safety analyses used the 316 mPDAC patients of NAPOLI-1 and
NCT02551991, one record each (the first grade >= 3 event, if any).

```{r population}
str(ui_pk$population)
```

# Source trace

Every value in the popPK model is in Table 2, which is identical to
Supplementary Table S3; the equations are from the final NONMEM control
stream printed in the Supplementary Material.

| Quantity | Value | Source |
|---|---|---|
| Structure: 2-cmt irinotecan, 1-cmt SN-38, direct + one-transit formation | -- | Figure 1; control stream `$MODEL`, `$PK` |
| `FM1 = FR1/(1+FR1+FR2)`, `FM2 = FR2/(1+FR1+FR2)` | -- | control stream `$PK` |
| SN-38 volume = irinotecan central volume | -- | Methods (Base model); control stream `VCM = VCP` |
| Scale: irinotecan `A1/V`, SN-38 `A2/(V/1000)` | -- | control stream `S1`, `S2` |
| `lcl` irinotecan CL | 17.9 L/week | Table 2 |
| `lvc` irinotecan central volume | 4.09 L | Table 2 |
| `lq` intercompartmental clearance | 1.35 L/week | Table 2 |
| `lvp` peripheral volume | 0.421 L | Table 2 |
| `lfr1` direct pathway ratio FR1 | 0.152 | Table 2, 'Fraction of direct ...' |
| `lfr2` delayed pathway ratio FR2 | 0.629 | Table 2, 'Fraction of delayed ...' |
| `lktr` transformation rate after delay | 2 1/week | Table 2 |
| `lcl_sn38` SN-38 clearance | 19,800 L/week | Table 2 |
| CL: Asian / previous site / female / oxaliplatin | x 1.204 / 1.515 / 0.799 / 1.339 | Table 2 |
| V: BSA exponent / previous site / female | (BSA/1.71)^0.573 / x 0.872 / 0.886 | Table 2 |
| FR2: previous site | x 1.376 | Table 2 |
| CL SN-38: bilirubin / CrCL / female / oxaliplatin | (BIL/0.41)^-0.266 / (CrCL/85.04)^0.25 / x 0.802 / 0.656 | Table 2 |
| IIV variances CL, V, FR1, FR2, KTR, CL SN-38 | 0.545, 0.066, 0.928, 0.188, 0.135, 0.126 | Table 2 |
| Covariances CL-V, CL-FR1, V-FR1 | 0.117, -0.558, -0.103 | Table 2 |
| Proportional error irinotecan / SN-38 | 0.243 / 0.291 | Table 2 |
| Irinotecan free-base molecular weight | 586.678 g/mol | Methods (Study data) |
| SN-38 molecular weight | 392.41 g/mol | not printed; computed from C22H20N2O5 |
| Diarrhea odds ratios (irinotecan / SN-38), per log10 Cavg,ss | 8.70 / 7.30 | Results, Diarrhea; Figure 4 |
| Neutropenia odds ratios (irinotecan / SN-38), per log10 Cavg,ss | 0.33 / 3.14 | Results, Neutropenia AEs; Figure 5 |
| Four logistic intercepts | -2.663, -1.575, -0.621, -1.165 | digitised from Figures 4a, 4b, 5a, 5b |

Two checks on the transcription: every printed %CV equals
`sqrt(exp(omega^2) - 1)` for its variance, and every printed correlation
equals the covariance divided by the root product of its two variances.

```{r trace-checks}
om <- c(cl = 0.545, vc = 0.066, fr1 = 0.928, fr2 = 0.188, ktr = 0.135, cl_sn38 = 0.126)
cv_printed <- c(85.2, 26.1, 124, 45.4, 38, 36.6)
cv_calc <- 100 * sqrt(exp(om) - 1)
rbind(printed = cv_printed, computed = round(cv_calc, 1))
stopifnot(max(abs(cv_calc - cv_printed) / cv_printed) < 0.005)

cor_calc <- c(
  cl_fr1 = -0.558 / sqrt(0.545 * 0.928),
  cl_vc = 0.117 / sqrt(0.545 * 0.066),
  vc_fr1 = -0.103 / sqrt(0.066 * 0.928)
)
rbind(printed = c(-0.785, 0.617, -0.416), computed = round(cor_calc, 3))
stopifnot(max(abs(cor_calc - c(-0.785, 0.617, -0.416))) < 0.002)

# The typical-value fractions of irinotecan elimination forming SN-38 match
# the Results ('one direct ... (9%) and one delayed via a transit
# compartment (35%)').
fm <- c(direct = 0.152, delayed = 0.629) / (1 + 0.152 + 0.629)
round(100 * fm, 1)
stopifnot(abs(100 * fm[["direct"]] - 9) < 0.5, abs(100 * fm[["delayed"]] - 35) < 0.5)
```

# Concentration units

The Results state that the joint model was developed with both analytes in
molar units, and the control stream's SN-38 scale `S2 = VCP/1000` turns
micromoles and litres into nmol/L. The paper's figures, however, are in
ug/mL (irinotecan) and ng/mL (SN-38). Because irinotecan is converted to
SN-38 one molecule for one molecule, only the SN-38 prediction depends on
which units the model was fitted in: read in mass units, it would be
586.678 / 392.41 = 1.50-fold higher than read in molar units. Every
parameter value is the same under both readings.

The model file follows the molar reading. It takes doses in mg of free base
and reports irinotecan in ug/mL, and it multiplies the SN-38 formation flux
by the molecular-weight ratio before reporting SN-38 in ng/mL. Besides the
Results statement, the paper's own model-derived exposures support this
choice. The median post hoc SN-38 Cavg,ss in Figures 4b and 5b sits at
log10 = -0.025; the molar reading reproduces it, and the mass reading lands
0.175 log10 units (50%) higher (see the steady-state section below). Total
irinotecan, where the two readings agree, reproduces the Figure 4a median,
which shows that the virtual cohort and dosing are right. The Figure 3 VPC
does not discriminate between the readings: its SN-38 to irinotecan ratios
fall between the two, closer to one in some bins and to the other in others.

# Virtual cohort

The cohort mirrors the exposure-safety population: NAPOLI-1 liposomal
irinotecan 100 mg/m^2 Q3W alone, 70 mg/m^2 Q2W with 5-FU/LV, and
NCT02551991 50 mg/m^2 Q2W with 5-FU/LV and oxaliplatin (NALIRIFOX). Arm
sizes keep the paper's roughly 260:56 split between NAPOLI-1 and
NCT02551991, which matters for the pooled comparison with Figure 4. All
patients receive product from the current manufacturing site. Covariates
are sampled to match Supplementary Table S1.

```{r cohort}
set.seed(20211215)
rxode2::rxSetSeed(20211215)

arms <- tibble::tribble(
  ~treatment,                      ~n,  ~dose_mgm2, ~tau, ~oxa,
  "NAPOLI-1 100 mg/m2 Q3W",        140, 100,         3,    0L,
  "NAPOLI-1 70 mg/m2 Q2W + 5-FU",  120,  70,         2,    0L,
  "NALIRIFOX 50 mg/m2 Q2W",         56,  50,         2,    1L
)
stopifnot(all(arms$n <= 200))

rtrunc <- function(n, mean, sd, lo, hi) pmin(pmax(rnorm(n, mean, sd), lo), hi)

cohort <- arms |>
  tidyr::uncount(n) |>
  mutate(
    id = dplyr::row_number(),
    BSA = rtrunc(dplyr::n(), 1.73, 0.22, 1.29, 2.48),
    SEXF = rbinom(dplyr::n(), 1, 0.486),
    RACE_ASIAN = rbinom(dplyr::n(), 1, 0.352),
    TBILI = 17.1 * exp(rtrunc(dplyr::n(), log(0.41), 0.4, log(0.12), log(2.11))),
    CRCL = rtrunc(dplyr::n(), 88, 30, 27, 177),
    CONMED_OXALIPLATIN = oxa,
    FORM_NALIRI_PREVSITE = 0L,
    amt = dose_mgm2 * BSA
  )
cohort |>
  summarise(
    n = dplyr::n(), `median BSA` = median(BSA), `% female` = 100 * mean(SEXF),
    `% Asian` = 100 * mean(RACE_ASIAN), `median bilirubin (mg/dL)` = median(TBILI / 17.1),
    `median CrCL (mL/min)` = median(CRCL), `% oxaliplatin` = 100 * mean(CONMED_OXALIPLATIN)
  ) |>
  knitr::kable(digits = 2)
```

# Simulation

Four cycles are simulated. Cycle 1 carries the profile for Figure 3; cycle 4
is at steady state and carries the Cavg,ss for Figures 4 and 5. Time is in
weeks. The model has two endpoints, so observation rows carry `dvid = 1`,
which returns both `Cc` and `Cc_sn38` on every row.

```{r events}
inf_dur <- 1.5 / 168 # 90-minute infusion, in weeks
grid_h <- c(0, 0.5, 1, 1.5, 2, 2.5, 3.5, 4.5, 6, 8, 12, 18, 24, 36, 48, 72, 96, 120, 168, 216, 264, 336, 408, 504)

make_events <- function(cohort) {
  doses <- cohort |>
    tidyr::crossing(cycle = 0:3) |>
    transmute(id, time = cycle * tau, amt, rate = amt / inf_dur, evid = 1L, cmt = "central", dvid = NA_integer_)
  obs <- cohort |>
    tidyr::crossing(cycle = c(0L, 3L), h = grid_h) |>
    filter(h <= tau * 168) |>
    transmute(id, time = cycle * tau + h / 168, amt = 0, rate = 0, evid = 0L, cmt = "central", dvid = 1L) |>
    distinct()
  covs <- cohort |> select(id, treatment, tau, BSA, SEXF, RACE_ASIAN, TBILI, CRCL, CONMED_OXALIPLATIN, FORM_NALIRI_PREVSITE)
  bind_rows(doses, obs) |>
    arrange(id, time, desc(evid)) |>
    left_join(covs, by = "id")
}
ev <- make_events(cohort)

sim <- as.data.frame(rxode2::rxSolve(
  mod_pk,
  events = ev,
  returnType = "data.frame",
  keep = c("treatment", "tau"),
  maxsteps = 1e6
))
stopifnot(!anyNA(sim$Cc), !anyNA(sim$Cc_sn38))
# LSODA undershoots zero by solver noise in the far tail; assert it is noise.
stopifnot(all(sim$Cc >= -1e-6 * max(sim$Cc)), all(sim$Cc_sn38 >= -1e-6 * max(sim$Cc_sn38)))
sim <- sim |> mutate(Cc = pmax(Cc, 0), Cc_sn38 = pmax(Cc_sn38, 0))
```

# Replicate Figure 3 (first-cycle profiles)

Figure 3 is a prediction-corrected VPC over the first week after a dose,
pooled over all seven studies and dose levels. The points below are the
simulated medians of Figure 3, digitised by the maintainers at each bin's
mid-time; they are a guide, not a target, because the cohort here is the
mPDAC subset only and the figure is prediction-corrected.

```{r fig3, fig.height = 7}
fig3 <- tibble::tribble(
  ~analyte,      ~time,  ~median,
  "irinotecan",  0.020,  29,
  "irinotecan",  0.095,  27,
  "irinotecan",  0.208,  12,
  "irinotecan",  0.333,  7.0,
  "irinotecan",  0.508,  4.7,
  "irinotecan",  0.804,  1.5,
  "SN-38",       0.020,  2.3,
  "SN-38",       0.095,  3.6,
  "SN-38",       0.208,  4.2,
  "SN-38",       0.333,  2.6,
  "SN-38",       0.508,  2.6,
  "SN-38",       0.804,  1.05
)

c1 <- sim |>
  filter(time > 0, time <= 1) |>
  select(id, time, `irinotecan` = Cc, `SN-38` = Cc_sn38) |>
  pivot_longer(c(`irinotecan`, `SN-38`), names_to = "analyte", values_to = "conc") |>
  group_by(analyte, time) |>
  summarise(
    p05 = quantile(conc, 0.05), p50 = median(conc), p95 = quantile(conc, 0.95),
    .groups = "drop"
  )

ggplot(c1, aes(time, p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.25) +
  geom_line() +
  geom_point(data = fig3, aes(time, median), colour = "red", size = 2) +
  scale_y_log10() +
  facet_wrap(~analyte, ncol = 1, scales = "free_y") +
  labs(
    x = "Time after dose (weeks)",
    y = "Concentration (irinotecan ug/mL; SN-38 ng/mL)",
    caption = "Replicates Figure 3 of Brendel 2021. Line and band: simulated median and 5th-95th percentiles.\nRed points: simulated medians of Figure 3, digitised."
  )
```

The SN-38 to irinotecan ratio at each bin cancels the dose and most
covariates, so it is the part of Figure 3 that bears on the unit reading. It
is not decisive: the published ratios lie between the molar and mass
readings, nearer the molar one in some bins and nearer the mass one in
others. The comparison uses the simulated
median at the grid time nearest each bin centre, and the figure is
prediction-corrected and pools all seven studies, so bin-level agreement is
only approximate.

```{r fig3-ratio}
ratio_tab <- fig3 |>
  pivot_wider(names_from = analyte, values_from = median) |>
  mutate(paper = `SN-38` / irinotecan) |>
  select(time, paper) |>
  left_join(
    c1 |>
      select(analyte, time, p50) |>
      pivot_wider(names_from = analyte, values_from = p50) |>
      mutate(time_bin = time) |>
      select(time_bin, irinotecan, `SN-38`),
    by = join_by(closest(time >= time_bin))
  ) |>
  mutate(
    `molar reading` = `SN-38` / irinotecan,
    `mass reading` = `molar reading` * 586.678 / 392.41
  ) |>
  select(`Time (weeks)` = time, `Figure 3 ratio` = paper, `molar reading`, `mass reading`)
knitr::kable(ratio_tab, digits = 3)
```

# Steady-state exposure and PKNCA

PKNCA computes each patient's average concentration over the last dosing
interval (cycle 4). With linear kinetics, Cavg,ss has closed forms,
`Dose / (CL * tau)` for irinotecan and
`1000 * MW ratio * (FM1 + FM2) * Dose / (CL_SN38 * tau)` for SN-38. PKNCA
must reproduce them.

```{r pknca}
ss <- sim |>
  filter(time >= 3 * tau - 1e-9) |>
  mutate(treatment = factor(treatment, levels = arms$treatment))
# Time-zero row of each steady-state interval is present by construction.
stopifnot(all(tapply(abs(ss$time - 3 * ss$tau) < 1e-9, ss$id, any)))

doses_ss <- ev |>
  filter(evid == 1, abs(time - 3 * tau) < 1e-9) |>
  mutate(treatment = factor(treatment, levels = arms$treatment)) |>
  select(id, time, amt, treatment)

intervals <- arms |>
  transmute(
    treatment = factor(treatment, levels = arms$treatment),
    start = 3 * tau, end = 4 * tau, cav = TRUE, cmax = TRUE, auclast = TRUE
  )

run_nca <- function(conc_col) {
  conc <- ss |>
    select(id, time, treatment, conc = all_of(conc_col)) |>
    filter(!is.na(conc))
  o_conc <- PKNCA::PKNCAconc(conc, conc ~ time | treatment + id)
  o_dose <- PKNCA::PKNCAdose(doses_ss, amt ~ time | treatment + id)
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose, intervals = as.data.frame(intervals)))
  as.data.frame(res$result)
}
nca_iri <- run_nca("Cc")
nca_sn38 <- run_nca("Cc_sn38")

indiv <- sim |>
  group_by(id) |>
  slice(1) |>
  ungroup() |>
  select(id, treatment, tau, cl, cl_sn38, fm1, fm2) |>
  left_join(cohort |> select(id, amt), by = "id") |>
  mutate(
    cav_iri_cf = amt / (cl * tau),
    cav_sn38_cf = 1000 * (392.41 / 586.678) * (fm1 + fm2) * amt / (cl_sn38 * tau)
  ) |>
  left_join(nca_iri |> filter(PPTESTCD == "cav") |> select(id, cav_iri = PPORRES), by = "id") |>
  left_join(nca_sn38 |> filter(PPTESTCD == "cav") |> select(id, cav_sn38 = PPORRES), by = "id")
stopifnot(nrow(indiv) == nrow(cohort), !anyNA(indiv$cav_iri), !anyNA(indiv$cav_sn38))

err <- indiv |>
  summarise(
    irinotecan_median = median(100 * (cav_iri / cav_iri_cf - 1)),
    irinotecan_p90 = quantile(abs(100 * (cav_iri / cav_iri_cf - 1)), 0.9),
    sn38_median = median(100 * (cav_sn38 / cav_sn38_cf - 1)),
    sn38_p90 = quantile(abs(100 * (cav_sn38 / cav_sn38_cf - 1)), 0.9)
  )
knitr::kable(err, digits = 2, caption = "PKNCA Cavg,ss versus the closed form (% difference)")
# The two sides use the same drawn parameters, so the difference is
# trapezoid error plus the small residual accumulation of slow-clearing
# patients after three doses. A mis-coded formation fraction, molecular-weight
# ratio or clearance moves these by tens of percent.
stopifnot(abs(err$irinotecan_median) < 2, err$irinotecan_p90 < 6)
stopifnot(abs(err$sn38_median) < 2, err$sn38_p90 < 6)
```

## Comparison with the published exposure distribution

The paper gives no NCA table; the only published exposure values are the
Cavg,ss quartiles of the 316 exposure-safety patients, drawn as vertical
lines in Figures 4 and 5. Their median is log10 0.50 (3.16 ug/mL) for total
irinotecan and log10 -0.025 (0.944 ng/mL) for SN-38.

```{r nca-compare}
sim_cav <- bind_rows(
  nca_iri |> filter(PPTESTCD == "cav") |> mutate(analyte = "Total irinotecan (ug/mL)"),
  nca_sn38 |> filter(PPTESTCD == "cav") |> mutate(analyte = "SN-38 (ng/mL)")
) |>
  select(analyte, PPTESTCD, PPORRES)
ref_cav <- data.frame(
  analyte = c("Total irinotecan (ug/mL)", "SN-38 (ng/mL)"),
  cav = 10^c(0.50, -0.025)
)
knitr::kable(ncaComparisonTable(sim_cav, ref_cav, by = "analyte"), digits = 3)

q_sim <- rbind(
  irinotecan = quantile(log10(indiv$cav_iri), c(0.25, 0.5, 0.75)),
  sn38_molar = quantile(log10(indiv$cav_sn38), c(0.25, 0.5, 0.75)),
  sn38_mass = quantile(log10(indiv$cav_sn38 * 586.678 / 392.41), c(0.25, 0.5, 0.75))
)
q_paper <- rbind(irinotecan = c(0.33, 0.50, 0.68), sn38 = c(-0.12, -0.025, 0.095))
knitr::kable(
  data.frame(
    Quantity = c("Total irinotecan", "SN-38, molar reading (model)", "SN-38, mass reading"),
    `Simulated Q1` = q_sim[, 1], `Simulated median` = q_sim[, 2], `Simulated Q3` = q_sim[, 3],
    `Figure 4 Q1` = q_paper[c(1, 2, 2), 1], `Figure 4 median` = q_paper[c(1, 2, 2), 2],
    `Figure 4 Q3` = q_paper[c(1, 2, 2), 3],
    check.names = FALSE
  ),
  digits = 3, row.names = FALSE,
  caption = "log10 Cavg,ss quartiles: simulated cohort against Figures 4 and 5"
)
# Median on the log10 scale: 0.1 is 26% on the concentration scale. The
# molar/mass readings differ by 0.175, so the gate discriminates them.
stopifnot(
  abs(q_sim["irinotecan", 2] - 0.50) < 0.1,
  abs(q_sim["sn38_molar", 2] - (-0.025)) < 0.1
)
```

The simulated interquartile ranges are wider than the published ones. That is
expected: the published quartiles come from post hoc (empirical Bayes)
estimates, which shrink towards the typical value, while the simulation draws
the full between-patient variability.

# Replicate Figures 4 and 5 (exposure-safety)

The four logistic models are univariable in log10 Cavg,ss. The figures plot
the fitted curve over each analyte's observed range, with the observed
proportion of patients with an event in each exposure quartile. Those
quartile points and their counts are printed in the figures and are overlaid
here.

```{r er-curves, fig.height = 6}
# The landmark models have no random effects, so rxode2 notes that a
# multi-subject solve has no omega; that note is expected and muffled.
muffle_no_omega <- function(w) {
  if (grepl("without 'omega'", conditionMessage(w), fixed = TRUE)) invokeRestart("muffleWarning")
}
solve_prob <- function(model, cav) {
  out <- withCallingHandlers(
    rxode2::rxSolve(
      model,
      events = data.frame(id = seq_along(cav), time = 0, amt = 0, evid = 0L, CAV = cav),
      returnType = "data.frame"
    ),
    warning = muffle_no_omega
  )
  out <- as.data.frame(out)
  pcol <- grep("^prob_", names(out), value = TRUE)
  stopifnot(length(pcol) == 1L, nrow(out) == length(cav))
  out[[pcol]]
}

ranges <- list(irinotecan = c(-0.76, 0.95), sn38 = c(-0.40, 0.58))
curves <- bind_rows(lapply(names(mod_er), function(nm) {
  an <- sub("^[a-z]+_", "", nm)
  x <- seq(ranges[[an]][1], ranges[[an]][2], length.out = 60)
  tibble::tibble(model = nm, log10_cav = x, p = solve_prob(mod_er[[nm]], 10^x))
}))

# Observed quartile points (x at the digitised point position; n and
# events as printed under each figure panel).
quart <- tibble::tribble(
  ~model,                    ~log10_cav, ~events, ~n,
  "diarrhea_irinotecan",      0.075,      6,      78,
  "diarrhea_irinotecan",      0.407,     12,      80,
  "diarrhea_irinotecan",      0.597,     19,      79,
  "diarrhea_irinotecan",      0.778,     19,      79,
  "diarrhea_sn38",           -0.194,     11,      79,
  "diarrhea_sn38",           -0.079,      9,      78,
  "diarrhea_sn38",            0.023,     17,      80,
  "diarrhea_sn38",            0.241,     19,      79,
  "neutropenia_irinotecan",   0.075,     30,      78,
  "neutropenia_irinotecan",   0.407,     18,      80,
  "neutropenia_irinotecan",   0.597,     14,      79,
  "neutropenia_irinotecan",   0.778,     16,      79,
  "neutropenia_sn38",        -0.194,     13,      79,
  "neutropenia_sn38",        -0.079,     21,      78,
  "neutropenia_sn38",         0.023,     20,      80,
  "neutropenia_sn38",         0.241,     24,      79
) |>
  mutate(observed = events / n)

ggplot(curves, aes(log10_cav, p)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(data = quart, aes(y = observed), colour = "orange", size = 2.5) +
  facet_wrap(~model, scales = "free_x") +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "log10 Cavg,ss (irinotecan ug/mL; SN-38 ng/mL)",
    y = "Probability of a grade >= 3 event",
    caption = "Replicates Figures 4 and 5 of Brendel 2021 (orange: observed quartile proportions)."
  )
```

## Intercept check

For a logistic regression fitted by maximum likelihood, the predicted
probabilities summed over the analysis set equal the number of events
exactly. Approximating each patient's exposure by the position of their
quartile's point, the digitised intercepts must therefore return close to
the printed totals: 56 of 316 diarrhea events (17.7%) and 78 of 316
neutropenia events (24.7%). This check does not use the digitised curves,
only the printed counts and the printed odds ratios.

```{r er-check}
chk <- quart |>
  group_by(model) |>
  group_modify(function(d, key) {
    p <- solve_prob(mod_er[[key$model]], 10^d$log10_cav)
    tibble::tibble(
      events_obs = sum(d$events), events_pred = sum(d$n * p), n = sum(d$n),
      max_bin_diff_pct = max(abs(100 * (p - d$observed)))
    )
  }) |>
  ungroup() |>
  mutate(
    pct_obs = 100 * events_obs / n, pct_pred = 100 * events_pred / n,
    diff_pct_points = pct_pred - pct_obs
  )
knitr::kable(chk, digits = 2)
stopifnot(all(chk$events_obs == c(56, 56, 78, 78)))
# The within-quartile spread of exposure makes the quartile-point
# approximation inexact; a misread intercept moves this by far more
# (0.5 on the logit scale shifts the diarrhea totals by about 7 points).
stopifnot(all(abs(chk$diff_pct_points) < 2))
```

The reported odds ratios can also be checked against the printed percentage
changes: an odds ratio of 8.70 per log10 unit means a doubling of total
irinotecan Cavg,ss raises the odds of grade >= 3 diarrhea by
`8.70^log10(2)` = `r round(8.70^log10(2), 2)`-fold.

## Predicted incidence in the virtual cohort

```{r er-cohort}
inc <- indiv |>
  mutate(
    diarrhea_irinotecan = solve_prob(mod_er$diarrhea_irinotecan, cav_iri),
    diarrhea_sn38 = solve_prob(mod_er$diarrhea_sn38, cav_sn38),
    neutropenia_irinotecan = solve_prob(mod_er$neutropenia_irinotecan, cav_iri),
    neutropenia_sn38 = solve_prob(mod_er$neutropenia_sn38, cav_sn38)
  ) |>
  group_by(treatment) |>
  summarise(across(diarrhea_irinotecan:neutropenia_sn38, \(x) 100 * mean(x)), .groups = "drop")
knitr::kable(inc, digits = 1, caption = "Mean predicted probability (%) of a grade >= 3 event by arm")
```

The paper reports 16.9% (NAPOLI-1) and 21.4% (NCT02551991) grade >= 3
diarrhea, and 20.8% and 42.9% grade >= 3 neutropenia AEs. The univariable
exposure models do not carry that difference between studies. The
irinotecan neutropenia model does predict more neutropenia on NALIRIFOX,
because the lower dose and the oxaliplatin-driven rise in irinotecan
clearance lower irinotecan exposure and the fitted relationship is inverse,
but it falls well short of the observed 42.9%. The authors call the inverse
irinotecan-neutropenia relationship unexpected, and the SN-38 neutropenia
relationship is not statistically significant (p = 0.115). These models
describe association in the pooled data; they are not a basis for predicting
the effect of adding oxaliplatin.

# Assumptions and deviations

* **Molar model, mass outputs.** Doses are mg of irinotecan free base; the
  SN-38 formation flux is multiplied by 392.41 / 586.678 so that SN-38 is
  reported in ng/mL. The SN-38 molecular weight is not in the paper; it was
  computed from the formula C22H20N2O5 with standard atomic weights. See
  "Concentration units" for why the molar reading was chosen.
* **Covariate coding.** Categorical effects are coded exactly as in the
  control stream, `(1 + THETA * indicator)`; the model file stores
  THETA = printed factor - 1. The control stream codes manufacturing site
  with the current site as `MFG = 1`, so `FORM_NALIRI_PREVSITE = 1 - MFG`.
  `SEX = 1` is female. The control stream sets the bilirubin and
  creatinine-clearance effects to 1 when either value is missing (`-99`);
  impute the medians (0.41 mg/dL, 85.04 mL/min) to reproduce that.
* **Bilirubin units.** The canonical `TBILI` column is in umol/L; the model
  divides by 17.1 to recover the source's mg/dL before applying
  `(BIL / 0.41)^-0.266`.
* **Creatinine clearance** is absolute mL/min, as in the source; the
  estimating equation is not reported.
* **Correlated residual error not encoded.** The control stream estimates a
  correlation of 0.323 between the irinotecan and SN-38 residuals by reusing
  `EPS(1)` on SN-38 records. Irinotecan and SN-38 are separate records in
  the data, so their residual draws are independent and the correlation
  cannot act; the SN-38 residual variance is `0.291^2` either way. nlmixr2
  has no cross-endpoint residual correlation, so only the two proportional
  errors are encoded.
* **Data below the limit of quantification** were discarded in the final
  model (M1), with 23% of irinotecan and 25% of SN-38 samples below the
  limit. This does not change the model, but explains why Figure 3's late
  observed medians sit above the simulated ones.
* **Exposure-safety intercepts are digitised.** The paper prints only the
  odds ratios and p values. The intercepts were recovered by tracing each
  fitted curve in Figures 4a, 4b, 5a and 5b (about 1000 pixel columns per
  panel) and fitting the intercept with the slope held at the printed odds
  ratio. Fitting both coefficients freely gives odds ratios of 8.83, 7.39,
  0.33 and 3.17 against the printed 8.70, 7.30, 0.33 and 3.14, which
  confirms the digitisation, and the intercept check above confirms that
  the digitised intercepts reproduce the printed event counts. The diarrhea
  odds ratios are described in the Results only as per "log-transformed"
  Cavg; the figure axes, and the free fits, show the logarithm is base 10.
* **Placeholder residual.** Each exposure-safety model carries a fixed
  additive residual of 0.001 on the probability so that rxode2 accepts an
  observation declaration; the source regressions are Bernoulli and have no
  residual.
* **C<sub>max</sub> analyses** (Figures S9, S11, S13 and S15) and the
  analyses at the first adverse event (Figures S6, S7) are not extracted;
  the paper reports neither odds ratios nor p values for them, and states
  that they gave the same conclusions.
* **Virtual cohort.** Covariate distributions are sampled independently
  from the Supplementary Table S1 summaries; correlations between
  covariates are not reproduced. The NAPOLI-1 split between the two
  liposomal-irinotecan arms is not reported and was set to 140:120.
