---
title: "Fluvoxamine: CYP2D6 phenotype and smoking (Britz 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fluvoxamine: CYP2D6 phenotype and smoking (Britz 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Britz 2019 is primarily a whole-body PBPK paper: it builds PK-Sim / MoBi models
of fluvoxamine and theophylline and couples them with existing caffeine,
rifampicin and midazolam models into a CYP1A2 drug-drug-interaction network.
Alongside that work the authors ran a **NONMEM population-PK analysis of
fluvoxamine**, reported in full in Supplement S1 section 4, to quantify the
effect of CYP2D6 poor metabolism and of cigarette smoking on fluvoxamine
clearance and to check those effect sizes against the PBPK predictions.

It is that population-PK analysis that is packaged here. The two source studies
were **analysed separately** (Supplement S1 section 4.2: "The two studies were
analyzed separately using ADVAN6"), each yielding its own set of parameter
estimates in Table S11, so the analysis is packaged as two independent model
files following the paper's own structure. The PK-Sim whole-body PBPK layer is
not reproduced: its tissue-plasma partition coefficients are reported only as a
calculation-method name ("Schmitt", "PK-Sim", "R+R") and its organ volumes and
blood flows are platform database outputs that appear nowhere in the paper.

```{r models}
mod_cyp2d6 <- readModelDb("Britz_2019_fluvoxamine_cyp2d6")
mod_smoking <- readModelDb("Britz_2019_fluvoxamine_smoking")
```

* Citation: `r rxode2::rxode2(mod_cyp2d6)$reference`
* Article: <https://doi.org/10.1002/psp4.12397>
* Supplement S1 (model documentation, Tables S1a-S11): distributed with the
  article on the CPT:PSP website and via Europe PMC
  (`PMC6539736`, supplementary file `psp412397-sup-0001-SupplementS1.pdf`).

## `Britz_2019_fluvoxamine_cyp2d6`

`r rxode2::rxode2(mod_cyp2d6)$description`

## `Britz_2019_fluvoxamine_smoking`

`r rxode2::rxode2(mod_smoking)$description`

# Population

Both models describe a **single 50 mg oral dose of fluvoxamine in healthy
volunteers**, with blood sampled pre-dose and at 1, 2, 3, 4, 5, 6, 7, 8, 10,
12, 24, 32 and 48 h; plasma concentrations were measured by HPLC with a lower
limit of quantification of 0.5 nmol/L (Supplement S1 section 4.2).

**Spigset 1997 (CYP2D6 model).** Ten young healthy volunteers categorised by
CYP2D6 phenotype: five extensive metabolizers (the single female plus four
males) and five poor metabolizers (all male). All subjects were non-smokers.
Mean age 25 years, body weight 55-86 kg. 139 concentrations entered the
analysis.

**Spigset 1995 (smoking model).** Twenty-four young healthy volunteers, 12
non-smokers and 12 smokers, with five females and seven males in each group.
All were CYP2D6 extensive metabolizers. Mean age 36.5 years, body weight
51-95 kg. 311 concentrations entered the analysis.

Because each study varies only one factor and holds the other fixed, the two
covariates are cleanly separated: the CYP2D6 model is estimated entirely in
non-smokers and the smoking model entirely in CYP2D6 extensive metabolizers.
Neither model can therefore speak to the combination of the two.

The same information is available programmatically from each model's
`population` metadata:

```{r population}
str(readModelDb("Britz_2019_fluvoxamine_cyp2d6")()$population, max.level = 1)
```

# Source trace

Every value below is transcribed from **Britz 2019 Supplement S1, Table S11**
("Parameter estimates of the final population pharmacokinetic models"), whose
two numeric columns are headed "Spigset 1997" and "Spigset 1995". The same
citations appear as in-file comments beside each `ini()` entry in
`inst/modeldb/specificDrugs/Britz_2019_fluvoxamine_cyp2d6.R` and
`inst/modeldb/specificDrugs/Britz_2019_fluvoxamine_smoking.R`.

| Equation / parameter | `_cyp2d6` (Spigset 1997) | `_smoking` (Spigset 1995) | Source location |
|---|---|---|---|
| `ld1` - zero-order input duration D | 1.53 h (RSE 12%) | 3.51 h (RSE 9%) | Table S11 row "D (h)"; corroborated in section 4.3 prose |
| `ltlag` - absorption lag time ALAG | 2.75 h (RSE 1%) | 1.79 h (RSE 6%) | Table S11 row "ALAG (h)"; corroborated in section 4.3 prose |
| `lvc` - apparent central volume Vc/F | 2610 L (RSE 11%) | 3030 L (RSE 12%) | Table S11 row "VCentral (l/F)"; corroborated in section 4.3 prose |
| `lcl` - apparent clearance CL/F | 147 L/h (RSE 25%) | 133 L/h (RSE 18%) | Table S11 row "CL (l/h/F)"; corroborated in section 4.3 prose |
| `e_cyp2d6_pm_cl` - CYP2D6 PM effect on CL/F | 0.775 (RSE 33%) | n/a | Table S11 row "CYP2D6 poor metabolism on CL" |
| `e_smoke_cl` - smoking effect on CL/F | n/a | 1.28 (RSE 21%) | Table S11 row "Smoking on CL" |
| `etalvc` - IIV on Vc/F | 29 %CV (RSE 21%) -> 0.0807501 | 53 %CV (RSE 18%) -> 0.2475630 | Table S11 row "IIV VCentral (%CV)"; `omega^2 = log(1 + CV^2)` |
| `etalcl` - IIV on CL/F | 49 %CV (RSE 17%) -> 0.2151920 | 49 %CV (RSE 22%) -> 0.2151920 | Table S11 row "IIV CL (%CV)"; `omega^2 = log(1 + CV^2)` |
| `propSd` - proportional residual SD | 0.34 (RSE 13%) | 0.49 (RSE 18%) | Table S11 row "Proportional (%)" |
| `addSd` - additive residual SD | `fixed(2.86506e-07)` ng/mL | 9.5502e-04 ng/mL (RSE 50%) | Table S11 row "Additive (nmol/ml)": (9e-10) fixed, and 3e-6; converted with MW 318.34 g/mol from Table S1b |
| Structure: 1 compartment, zero-order input, lag, linear elimination | n/a | n/a | Section 4.3 "Population pharmacokinetic model"; section 4.2 model-building narrative |
| Exponential (log-normal) IIV | n/a | n/a | Section 4.2: "IIVs were modelled exponentially" |
| Combined proportional + additive residual error | n/a | n/a | Section 4.3: "Residual variability was best described with a combined error model" |
| MW 318.34 g/mol (unit conversion only) | n/a | n/a | Supplement S1 Table S1b row "MW" |

Reference values used later for validation come from **Supplement S1 Table S1d**
("Observed and predicted AUC and Cmax values of fluvoxamine"), rows for
Spigset 1995 and Spigset 1997 at 50 mg p.o. single dose.

# Virtual cohort

The original subject-level data are not public, so the figures below use virtual
cohorts of 200 subjects per arm. Only the covariate that each study varied is
set; no body-weight or age covariate enters either model, so nothing else needs
to be sampled.

Observation times combine a 0.25 h grid over the absorption phase with an hourly
grid to 72 h, plus the exact end-of-input time `tlag + d1` for each model so
that Cmax and Tmax are resolved exactly rather than to the nearest grid point.

```{r cohort}
dose_mg <- 50

# End of the zero-order input = tlag + d1, the exact time of the peak.
peak_time <- c(cyp2d6 = 2.75 + 1.53, smoking = 1.79 + 3.51)

obs_grid <- function(peak) {
  sort(unique(c(seq(0, 12, by = 0.25), seq(12, 72, by = 1), peak)))
}

# One arm's event table. `rate = -2` is mandatory: it tells rxode2 to use the
# modelled duration dur(central) = d1. Without it the dose collapses to an
# instantaneous bolus.
make_arm <- function(n, cov_name, cov_value, arm, peak, id_offset = 0L) {
  sched <-
    rxode2::et(amt = dose_mg, cmt = "central", rate = -2) |>
    rxode2::et(obs_grid(peak), cmt = "central") |>
    as.data.frame()
  out <- do.call(rbind, lapply(seq_len(n), function(i) {
    x <- sched
    x$id <- id_offset + i
    x
  }))
  out[[cov_name]] <- cov_value
  out$arm <- arm
  out
}

n_per_arm <- 200L

ev_cyp2d6 <- dplyr::bind_rows(
  make_arm(n_per_arm, "CYP2D6_PM", 0L, "CYP2D6 EM", peak_time[["cyp2d6"]], 0L),
  make_arm(n_per_arm, "CYP2D6_PM", 1L, "CYP2D6 PM", peak_time[["cyp2d6"]], n_per_arm)
)
ev_smoking <- dplyr::bind_rows(
  make_arm(n_per_arm, "SMOKE", 0L, "Non-smoker", peak_time[["smoking"]], 0L),
  make_arm(n_per_arm, "SMOKE", 1L, "Smoker", peak_time[["smoking"]], n_per_arm)
)

# Disjoint ids per arm: duplicate ids would silently merge into one subject
# receiving the summed dose.
stopifnot(
  !anyDuplicated(unique(ev_cyp2d6[, c("id", "time", "evid")])),
  !anyDuplicated(unique(ev_smoking[, c("id", "time", "evid")]))
)
```

# Simulation

Two simulations are run for each model: a stochastic cohort that carries the
published inter-individual variability (used for the prediction-interval
figures) and a typical-value solve with the random effects zeroed (used for the
NCA comparison and the structural checks, both of which must be exactly
reproducible).

```{r simulate}
rxode2::rxSetSeed(20190501)
sim_cyp2d6 <- rxode2::rxSolve(mod_cyp2d6, ev_cyp2d6, keep = "arm") |>
  as.data.frame()

rxode2::rxSetSeed(20190502)
sim_smoking <- rxode2::rxSolve(mod_smoking, ev_smoking, keep = "arm") |>
  as.data.frame()
```

```{r simulate-typical}
typ_arm <- function(cov_name, cov_value, arm, peak, id_offset) {
  make_arm(1L, cov_name, cov_value, arm, peak, id_offset)
}

ev_typ_cyp2d6 <- dplyr::bind_rows(
  typ_arm("CYP2D6_PM", 0L, "CYP2D6 EM", peak_time[["cyp2d6"]], 1L),
  typ_arm("CYP2D6_PM", 1L, "CYP2D6 PM", peak_time[["cyp2d6"]], 2L)
)
ev_typ_smoking <- dplyr::bind_rows(
  typ_arm("SMOKE", 0L, "Non-smoker", peak_time[["smoking"]], 1L),
  typ_arm("SMOKE", 1L, "Smoker", peak_time[["smoking"]], 2L)
)

typ_cyp2d6 <- rxode2::rxSolve(
  rxode2::zeroRe(mod_cyp2d6), ev_typ_cyp2d6, keep = "arm"
) |> as.data.frame()
typ_smoking <- rxode2::rxSolve(
  rxode2::zeroRe(mod_smoking), ev_typ_smoking, keep = "arm"
) |> as.data.frame()
```

# Replicate published figures

## Figure S11c - impact of CYP2D6 phenotype and smoking

Supplement S1 Figure S11c plots simulated typical plasma concentration-time
profiles for a single 50 mg oral dose from the two final population-PK models:
CYP2D6 extensive versus poor metabolizers in the left panel, and non-smokers
versus smokers in the right panel. The chunk below reproduces both panels.

```{r figure-S11c}
# Replicates Figure S11c of Britz 2019 Supplement S1.
dplyr::bind_rows(
  typ_cyp2d6 |> dplyr::mutate(panel = "CYP2D6 phenotype (Spigset 1997)"),
  typ_smoking |> dplyr::mutate(panel = "Smoking status (Spigset 1995)")
) |>
  dplyr::filter(!is.na(Cc)) |>
  ggplot(aes(time, Cc, colour = arm, linetype = arm)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~panel) +
  coord_cartesian(xlim = c(0, 48)) +
  labs(
    x = "Time (h)", y = "Fluvoxamine concentration (ng/mL)",
    colour = NULL, linetype = NULL,
    title = "Figure S11c - typical profiles after a single 50 mg oral dose",
    caption = "Replicates Figure S11c of Britz 2019 Supplement S1."
  ) +
  theme(legend.position = "bottom")
```

The two published qualitative findings are reproduced: poor metabolizers show
the higher and longer-lasting exposure of the two CYP2D6 strata, and smokers
show the lower exposure of the two smoking strata. In both panels the curves
are indistinguishable during the absorption phase and separate only after the
peak, because each covariate acts on clearance alone.

## Prediction intervals across the published variability

```{r figure-vpc}
dplyr::bind_rows(
  sim_cyp2d6 |> dplyr::mutate(panel = "CYP2D6 phenotype (Spigset 1997)"),
  sim_smoking |> dplyr::mutate(panel = "Smoking status (Spigset 1995)")
) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(panel, arm, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.18, colour = NA) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~panel) +
  coord_cartesian(xlim = c(0, 48)) +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Fluvoxamine concentration (ng/mL)",
    colour = NULL, fill = NULL,
    title = "Median and 5th-95th percentile prediction intervals",
    caption = paste(
      "Virtual cohorts of", n_per_arm,
      "subjects per arm carrying the Table S11 inter-individual variability."
    )
  ) +
  theme(legend.position = "bottom")
```

# Structural checks

Because the models are one-compartment and linear, several quantities have exact
closed forms. Both sides of each comparison below use the same parameter values,
so the only difference is ODE-solver and NCA numerical error and a tight bound
is appropriate.

```{r structural-checks}
p_cyp2d6 <- rxode2::rxode2(mod_cyp2d6)$theta
p_smoking <- rxode2::rxode2(mod_smoking)$theta

expected <- tibble::tibble(
  arm = c("CYP2D6 EM", "CYP2D6 PM", "Non-smoker", "Smoker"),
  cl = c(
    exp(p_cyp2d6[["lcl"]]),
    exp(p_cyp2d6[["lcl"]]) * p_cyp2d6[["e_cyp2d6_pm_cl"]],
    exp(p_smoking[["lcl"]]),
    exp(p_smoking[["lcl"]]) * p_smoking[["e_smoke_cl"]]
  ),
  vc = c(
    exp(p_cyp2d6[["lvc"]]), exp(p_cyp2d6[["lvc"]]),
    exp(p_smoking[["lvc"]]), exp(p_smoking[["lvc"]])
  ),
  tlag_plus_d1 = c(
    rep(exp(p_cyp2d6[["ltlag"]]) + exp(p_cyp2d6[["ld1"]]), 2),
    rep(exp(p_smoking[["ltlag"]]) + exp(p_smoking[["ld1"]]), 2)
  )
) |>
  dplyr::mutate(
    # AUC0-inf = Dose / (CL/F); dose in mg, CL in L/h, Cc reported in ng/mL.
    aucinf_expected = dose_mg / cl * 1000,
    half_life_expected = log(2) * vc / cl,
    tmax_expected = tlag_plus_d1
  )

knitr::kable(
  expected |>
    dplyr::select(arm, cl, vc, tmax_expected, half_life_expected, aucinf_expected) |>
    dplyr::rename(
      "Arm" = arm,
      "CL/F (L/h)" = cl,
      "Vc/F (L)" = vc,
      "Expected Tmax (h)" = tmax_expected,
      "Expected t1/2 (h)" = half_life_expected,
      "Expected AUC0-inf (ng*h/mL)" = aucinf_expected
    ),
  digits = 3,
  caption = "Closed-form expectations implied by the Table S11 estimates."
)
```

```{r structural-gates}
typ_all <- dplyr::bind_rows(typ_cyp2d6, typ_smoking) |>
  dplyr::filter(!is.na(Cc))

observed_peaks <- typ_all |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    cmax = max(Cc),
    tmax = time[which.max(Cc)],
    .groups = "drop"
  ) |>
  dplyr::left_join(expected, by = "arm")

# Tmax must equal tlag + d1 exactly: the zero-order input stops there and
# elimination takes over.
stopifnot(all(abs(observed_peaks$tmax - observed_peaks$tmax_expected) < 1e-6))

# The covariate effects are pure multipliers on CL/F, so the exposure ratios
# invert the Table S11 multipliers exactly.
auc_ratio <- function(df, num, den) {
  a <- function(k) {
    d <- dplyr::filter(df, arm == k)
    sum(diff(d$time) * (head(d$Cc, -1) + tail(d$Cc, -1)) / 2)
  }
  a(num) / a(den)
}
# Truncated AUCs are used here, so allow a little room for the different
# extrapolated tails; the exact test on AUC0-inf is in the NCA section below.
stopifnot(
  abs(auc_ratio(typ_cyp2d6, "CYP2D6 PM", "CYP2D6 EM") - 1 / 0.775) < 0.05,
  abs(auc_ratio(typ_smoking, "Smoker", "Non-smoker") - 1 / 1.28) < 0.05
)
```

# PKNCA validation

NCA is run on the typical-value profiles, which makes the result exactly
reproducible and directly comparable with the closed forms above.

```{r pknca}
nca_for <- function(sim, events) {
  conc <- sim |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::select(id, time, Cc, arm)
  dose <- events |>
    dplyr::filter(evid == 1) |>
    dplyr::select(id, time, amt, arm)
  PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(conc, Cc ~ time | arm + id),
    PKNCA::PKNCAdose(dose, amt ~ time | arm + id),
    intervals = data.frame(
      start = 0, end = Inf,
      cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
    )
  ))
}

nca_cyp2d6 <- nca_for(typ_cyp2d6, ev_typ_cyp2d6)
nca_smoking <- nca_for(typ_smoking, ev_typ_smoking)

nca_all <- dplyr::bind_rows(
  as.data.frame(nca_cyp2d6),
  as.data.frame(nca_smoking)
) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  dplyr::select(arm, PPTESTCD, PPORRES)
```

The NCA output reproduces every closed form to within numerical error, which
confirms that the packaged models carry the Table S11 values the paper printed:

```{r pknca-gate}
check <- nca_all |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::left_join(expected, by = "arm")

stopifnot(
  all(abs(check$aucinf.obs / check$aucinf_expected - 1) < 0.002),
  all(abs(check$half.life / check$half_life_expected - 1) < 0.002),
  all(abs(check$tmax - check$tmax_expected) < 1e-6)
)

knitr::kable(
  check |>
    dplyr::select(arm, tmax, cmax, half.life, aucinf.obs) |>
    dplyr::rename(
      "Arm" = arm,
      "Tmax (h)" = tmax,
      "Cmax (ng/mL)" = cmax,
      "t1/2 (h)" = half.life,
      "AUC0-inf (ng*h/mL)" = aucinf.obs
    ),
  digits = 2,
  caption = "Typical-value NCA from the packaged models."
)
```

## Comparison against published NCA

Supplement S1 Table S1d reports the **observed** AUC0-inf and Cmax of the two
Spigset studies at 50 mg p.o., separately for each covariate stratum. Those
observed values are the data the population-PK models were fitted to, so they
are the right reference for this comparison.

```{r nca-compare}
# Britz 2019 Supplement S1 Table S1d, rows "50 po (tab), s.d." for
# Spigset 1997 (CYP2D6 EM n=10, PM n=5) and Spigset 1995 (non-smokers n=12,
# smokers n=12). AUC values are AUC0-inf.
published <- tibble::tribble(
  ~arm,         ~cmax,  ~aucinf.obs,
  "CYP2D6 EM",  14.20,  318.30,
  "CYP2D6 PM",  16.00,  417.00,
  "Non-smoker", 18.40,  353.36,
  "Smoker",     12.45,  245.44
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_all,
  reference = published,
  by = "arm",
  units = c(cmax = "ng/mL", aucinf.obs = "ng*h/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Typical-value NCA vs. the observed values in Britz 2019 Supplement S1 Table S1d. * differs from the reference by more than 20%.",
  align = c("l", "l", "r", "r", "r")
)
attr(cmp, "footnote")
```

**AUC0-inf** agrees closely in every arm: the typical-value AUC is
`Dose / (CL/F)` by construction, and the three non-smoker arms land within 7% of
the observed means. The smoker arm is the weakest at about +20%, which is also
the arm where the paper's own PBPK model performed worst on AUC
(Table S1d gives a PBPK predicted/observed AUC ratio of 1.04 for smokers against
1.18-1.49 for the other three arms, but its smoker prediction was tuned by the
optimised 1.38-fold CYP1A2 induction factor, whereas the population-PK model
estimates a single clearance multiplier of 1.28 from the same data).

**Cmax** is reproduced less well, and the CYP2D6 EM arm is flagged. Two
structural features of the model explain this and neither is a transcription
error:

* Table S11 estimates **no inter-individual variability on `D` or `ALAG`**, so
  every simulated subject reaches its peak at exactly the same time. Real
  subjects do not, so averaging observed profiles across subjects with differing
  individual Tmax flattens and lowers the observed mean peak relative to any
  single typical profile. The models were fitted by FOCE-I to the whole
  concentration-time profile, not to Cmax.
* Because both covariates act on clearance only, and because the absorption
  phase is long relative to the half-life, Cmax is nearly insensitive to the
  covariate: the model predicts a CYP2D6 PM/EM Cmax ratio of about 1.01 against
  an observed 1.13. The covariate's effect lives in the terminal phase and hence
  in AUC, which the model does reproduce.

The paper's own PBPK model shows the same direction of error on the flagged
arm, predicting a Cmax of 16.53 ng/mL against the observed 14.20 ng/mL for
Spigset 1997 extensive metabolizers (Table S1d). No parameter has been adjusted
to improve any of these comparisons.

# Assumptions and deviations

* **Only the population-PK layer of the paper is packaged.** Britz 2019 is
  mainly a PK-Sim / MoBi whole-body PBPK study of fluvoxamine and theophylline
  embedded in a five-drug CYP1A2 DDI network. That PBPK layer is not
  reproducible from the publication: Supplement S1 Tables S1b and S2b give the
  organ-plasma partition coefficients only as a calculation-method name
  ("Schmitt", "PK-Sim", "R+R") with no values, and no organ volume, blood flow
  or tissue composition appears anywhere in the paper or supplement. Those are
  PK-Sim database outputs. The drug-dependent parameters that *are* published
  (MW 318.34 g/mol, logP 3.57, fu 23%, CYP1A2 KM 7.35 nmol/L and kcat
  0.016/min non-smoker and 0.022/min smoker, CYP2D6 KM 76.30 umol/L and kcat
  110.56/min for extensive metabolizers and 0 for poor metabolizers) are
  recorded here for provenance but are not sufficient to rebuild the model.
* **Two model files, one per study.** Supplement S1 section 4.2 states that
  "The two studies were analyzed separately using ADVAN6", and Table S11 reports
  two independent parameter columns. The two fits are therefore packaged as two
  separate model files rather than as one model with two covariates. Neither
  model was fitted to data varying both factors, so combining the CYP2D6 and
  smoking multipliers on a single clearance would go beyond what the paper
  supports.
* **IIV scale.** Table S11 reports the inter-individual variability as a
  percent coefficient of variation and section 4.2 states that IIVs were
  "modelled exponentially", so the variances are computed as
  `omega^2 = log(1 + CV^2)`. If the authors instead used the common NONMEM
  shorthand `%CV = 100 * omega`, the variances would be up to 13% larger
  (for example 0.2809 instead of 0.2476 at 53 %CV); that is well inside the
  18-22% relative standard errors Table S11 reports for those omegas, and no
  control stream was published that could settle the convention. The supplement
  ships only PK-Sim project files (SQLite `.pksim5` archives), not the NONMEM
  control streams.
* **Additive residual error units.** Table S11 labels the additive residual
  error "nmol/ml" while section 4.2 reports the assay's limit of quantification
  as 0.5 nmol/l. The printed label is taken at face value and converted to the
  ng/mL scale used by `Cc` with the MW of 318.34 g/mol from Table S1b. The
  choice is immaterial: under either reading the additive term is below
  0.001 ng/mL against a Cmax near 15 ng/mL and a proportional error of 34-49%,
  so it contributes less than 0.01% of the residual. The paper says as much
  ("Although the additive error is very low, it was necessary to adequately
  describe the data").
* **The additive error of the CYP2D6 model was fixed, not estimated.**
  Table S11 prints it in parentheses and the table footnote states "Parameter
  values in parentheses were fixed", so it is wrapped in `fixed()`.
* **No inter-individual variability on `D`, `ALAG` or the covariate effects.**
  Table S11 reports omegas only for Vc/F and CL/F, so the packaged models carry
  only those two etas. No off-diagonal covariance is reported, so the etas are
  diagonal.
* **No covariate other than the study's own factor.** Neither model carries
  body weight, age or sex. The paper tested CYP2D6 phenotype as a continuous
  covariate (the dextromethorphan metabolic ratio, linear and logarithmic) as
  well as categorically and retained the categorical form; body weight was not
  among the covariates reported as tested.
* **Dosing.** Both models must be dosed into `central` with `rate = -2` so that
  rxode2 applies the modelled zero-order duration `dur(central) = d1`. A dose
  record without `rate = -2` collapses to an instantaneous bolus and produces a
  peak at the lag time with a substantially higher Cmax.
* **Virtual cohorts.** The figures use 200 subjects per arm rather than the
  studies' 5-12 subjects per stratum, so the prediction intervals reflect the
  published variability rather than the small-sample scatter of the original
  trials. The NCA comparison uses typical-value profiles, not cohort summaries,
  so it does not depend on the cohort size or on the random-number stream.
