---
title: "Midazolam preterm-neonate PBPK (Brussee 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Midazolam preterm-neonate PBPK (Brussee 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Brussee_2018_midazolam_pbpk"))$reference`
* Description: `r rxode2::rxode(readModelDb("Brussee_2018_midazolam_pbpk"))$description`
* Article: <https://doi.org/10.1002/psp4.12295>

# Population

Brussee 2018 fit a semi-physiological population PK model to pooled IV and
oral midazolam concentration data from 37 preterm neonates admitted to the
neonatal intensive care unit of the Sophia's Children Hospital (Rotterdam, NL).
The cohort had gestational age at birth 26-34 weeks, birth weight 745-2,135 g,
postnatal age 3-11 days, and body weight at the time of dosing 770-2,030 g.
The source data combine de Wildt 2001 (IV cohort, *Clin Pharmacol Ther* 70:525-531)
and de Wildt 2002 (oral cohort, *Br J Clin Pharmacol* 53:390-392): a 0.1 mg/kg
dose given either orally via a nasogastric tube (n = 13) or as a 30-minute
intravenous infusion (n = 25), with a crossover dose by the alternate route
after >= 72 h in subjects still meeting inclusion criteria. Plasma midazolam
and 1-OH-midazolam concentrations were collected at 0.5, 1, 2, 4, 6, 12, and
24 h post-dose; below-LLOQ samples (< 1% of 329 midazolam and < 2% of 153
1-OH-midazolam observations) were discarded before fitting. No covariates
beyond body weight (which scales the physiological tissue volumes, blood
flows, and intestinal-length terms) reached statistical significance in the
covariate analysis.

The same information is available programmatically via
`readModelDb("Brussee_2018_midazolam_pbpk")$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to every
`ini()` entry in `inst/modeldb/specificDrugs/Brussee_2018_midazolam_pbpk.R`.
The table below collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Hepatic extraction `E_h = (f_u,B * CL_int,H) / (f_u,B * CL_int,H + Q_h)` | derived | Brussee 2018 Eq. 1 |
| Gut wall extraction `E_g = (f_u,G * CL_int,G) / (f_u,G * CL_int,G + Q_gut)` | derived | Brussee 2018 Eq. 2 |
| Qgut hybrid `Q_gut = (Q_villi * CL_perm) / (Q_villi + CL_perm)` | derived | Brussee 2018 Eq. 3 |
| Intestinal surface `A = 2*pi*radius*h`, radius `= 1 cm`, `h = 2.736 * WT[g]^0.512 cm` | derived | Brussee 2018 Eq. 4 (Ives 2016, Struijs 2009) |
| Oral bioavailability `F_total = F_a * F_g * F_h` | derived | Brussee 2018 Eq. 5 |
| Reference WT for term-neonate scaling | 3.55 kg | Bjorkman 2005 / Brussee 2018 Table 1 |
| Reference liver volume `V_h,3.55kg` | 0.120 L | Brussee 2018 Table 1 (Bjorkman 2005 ref 7) |
| Portal vein / liver volume ratio | 0.778 | Brussee 2018 Table 1 (Aguirre-Reyes 2015 ref 24) |
| Reference gut wall volume `V_gw,3.55kg` | 0.050 L | Brussee 2018 Table 1 |
| Reference hepatic blood flow `Q_h,3.55kg` | 13.2 L/h | Brussee 2018 Table 1 |
| Blood-flow ratios (Q_pv, Q_ha, Q_in, Q_muc, Q_villi) | 0.75, 0.25, 0.4, 0.8, 0.6 | Brussee 2018 Table 1 (Johnson 2006, Yang 2007 refs 5, 25, 32) |
| Effective intestinal permeability `P_eff,man` (midazolam) | 4.4e-4 cm/s = 1.584 cm/h | Brussee 2018 Table 1 (Yang 2007 ref 25) |
| Midazolam unbound fraction in blood `f_u,B` | 0.04094 | Brussee 2018 Table 1 (McNamara-Alcorn from f_u,adult = 0.0303, [P]_pediatric = 27.1, [P]_adult = 37.0) |
| Midazolam unbound fraction in gut `f_u,G` | 1 | Brussee 2018 Methods (Yang 2007 ref 25) |
| Midazolam blood:plasma `B:P` | 0.568 | Brussee 2018 Table 1 (Maharaj 2013 ref 30) |
| 1-OH-midazolam unbound in blood `f_u,M` | 0.1394 | Brussee 2018 Table 1 (from f_u,M,adult = 0.106, Mandema 1992 ref 29) |
| 1-OH-midazolam blood:plasma `B:P` | 0.613 | Brussee 2018 Table 1 |
| Fraction metabolised to 1-OH-midazolam `f_M` | 1 | Brussee 2018 Table 1 |
| Absorption rate constant `k_a` | 10 1/h (fixed) | Brussee 2018 Methods (Table 1; sensitivity analysis 4.16-25 1/h) |
| Midazolam intrinsic hepatic CL `CL_H,int` | 6.7 L/h (RSE 10%) | Brussee 2018 Table 2 |
| Midazolam intrinsic gut wall CL `CL_G,int` | 0.0196 L/h (RSE 178%) | Brussee 2018 Table 2 |
| Midazolam apparent V (blood) | 3.0 L (RSE 11%) | Brussee 2018 Table 2 |
| 1-OH-midazolam intrinsic hepatic CL `CL_H,int,M` | 8.9 L/h (RSE 22%) | Brussee 2018 Table 2 |
| 1-OH-midazolam apparent V (blood) | 2.7 L (RSE 43%) | Brussee 2018 Table 2 |
| IIV omega^2 (CL_H,int, V, CL_H,int,M, V_M) | 0.887, 0.603, 0.832, 0.887 | Brussee 2018 Table 2 |
| Proportional residual error sigma^2 (parent / metabolite) | 0.201 / 0.164 | Brussee 2018 Table 2 |
| Additive residual error sigma^2 (parent / metabolite) | 0.0001 FIX (both) | Brussee 2018 Table 2 |

# Virtual cohort

Original observed concentrations from de Wildt 2001/2002 are not openly
available. The figures below use a 200-subject virtual cohort whose body-weight
distribution approximates the published trial demographics: a uniform sample
from 0.77-2.03 kg (the body-weight range at the time of dosing reported in
Methods Data).

```{r cohort}
set.seed(20180510)  # paper's published-online date

n_sub <- 200L
wt_range <- c(0.77, 2.03)  # kg

cohort_wt <- tibble::tibble(
  id = seq_len(n_sub),
  WT = runif(n_sub, min = wt_range[1], max = wt_range[2])
)

# Observation grid. Brussee 2018 Methods sampled at 0.5, 1, 2, 4, 6, 12, 24 h;
# use a denser grid here for smoother concentration-time profiles.
obs_times <- c(0.01, seq(0.25, 24, by = 0.25))

make_events <- function(cohort, route, dose_per_kg = 0.1, infusion_h = 0.5) {
  doses <- cohort |>
    dplyr::mutate(
      EVID = 1L,
      AMT  = dose_per_kg * WT * 1000,         # mg/kg * kg * 1000 = ug
      TIME = 0,
      CMT  = ifelse(route == "oral", "depot", "central"),
      RATE = ifelse(route == "iv30",
                    dose_per_kg * WT * 1000 / infusion_h,
                    NA_real_),
      route = route
    ) |>
    dplyr::select(id, EVID, AMT, RATE, TIME, CMT, WT, route)

  obs <- tidyr::expand_grid(
    cohort,
    TIME = obs_times
  ) |>
    dplyr::mutate(EVID = 0L, AMT = 0, RATE = NA_real_,
                  CMT = "Cc", route = route) |>
    dplyr::select(id, EVID, AMT, RATE, TIME, CMT, WT, route)

  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, TIME, dplyr::desc(EVID))
}

events_oral <- make_events(cohort_wt, "oral")
events_iv   <- make_events(cohort_wt, "iv30")
```

# Simulation

```{r simulate}
mod <- readModelDb("Brussee_2018_midazolam_pbpk")

sim_oral <- rxode2::rxSolve(
  mod, events = events_oral,
  keep = c("route", "WT"),
  returnType = "data.frame"
) |>
  dplyr::mutate(route = "Oral")

sim_iv <- rxode2::rxSolve(
  mod, events = events_iv,
  keep = c("route", "WT"),
  returnType = "data.frame"
) |>
  dplyr::mutate(route = "IV 30-min infusion")

sim <- dplyr::bind_rows(sim_oral, sim_iv)
```

The typical-value (population-mean) profile, useful for Figure 4 replication,
zeros out the between-subject variability:

```{r simulate-typical}
mod_typical <- mod |> rxode2::zeroRe()

# Typical 1.1 kg neonate
typical_cohort <- tibble::tibble(id = 1L, WT = 1.1)
typical_oral <- rxode2::rxSolve(
  mod_typical, events = make_events(typical_cohort, "oral"),
  keep = c("route", "WT"),
  returnType = "data.frame"
) |> dplyr::mutate(route = "Oral")
typical_iv <- rxode2::rxSolve(
  mod_typical, events = make_events(typical_cohort, "iv30"),
  keep = c("route", "WT"),
  returnType = "data.frame"
) |> dplyr::mutate(route = "IV 30-min infusion")

typical <- dplyr::bind_rows(typical_oral, typical_iv)
```

# Replicate published figures

## Figure 4a -- Midazolam plasma profile (oral vs IV)

Brussee 2018 Figure 4a shows model-based simulations of midazolam plasma
concentrations following a 0.1 mg/kg dose given orally (black) or as a 30-min
infusion (gray) in the 37 preterm neonates from the study, with the
median and minimal / maximal envelope.

```{r figure-4a-midazolam, fig.cap = "Replicates Figure 4a of Brussee 2018."}
sim |>
  dplyr::filter(time > 0) |>
  dplyr::group_by(route, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50, colour = route, fill = route)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(
    x = "Time after dose (hours)",
    y = "Midazolam plasma concentration (ng/mL)",
    colour = NULL, fill = NULL,
    title = "Figure 4a -- midazolam, oral vs IV 30-min infusion",
    caption = "Median with 5th-95th percentile band, 200-subject virtual cohort (WT 0.77-2.03 kg)."
  )
```

## Figure 4b -- 1-OH-midazolam plasma profile (oral vs IV)

Brussee 2018 Figure 4b shows the same simulation for the 1-OH-midazolam
metabolite, where the oral curve climbs higher than the IV curve over the
first 4 h because of presystemic conversion in the gut wall and liver.

```{r figure-4b-metabolite, fig.cap = "Replicates Figure 4b of Brussee 2018."}
sim |>
  dplyr::filter(time > 0) |>
  dplyr::group_by(route, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc_1ohm, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc_1ohm, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc_1ohm, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50, colour = route, fill = route)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(
    x = "Time after dose (hours)",
    y = "1-OH-midazolam plasma concentration (ng/mL)",
    colour = NULL, fill = NULL,
    title = "Figure 4b -- 1-OH-midazolam, oral vs IV 30-min infusion",
    caption = "Median with 5th-95th percentile band; presystemic metabolism elevates the oral curve over 0-4 h."
  )
```

## Figure 2 / 3 -- extraction ratios and bioavailability vs body weight

Brussee 2018 Figures 2c and 3 report extraction ratios and bioavailability
as scatter plots of typical-value model predictions across body weight. The
extraction-ratio and bioavailability terms are computed inside `model()` and
are exposed as derived outputs by `rxSolve()`, so the relationship across
the cohort can be reproduced directly.

```{r figure-2c-3, fig.cap = "Replicates Figures 2c and 3 of Brussee 2018: extraction ratio and bioavailability vs body weight."}
typical_grid <- tibble::tibble(
  id = seq_len(50),
  WT = seq(min(wt_range), max(wt_range), length.out = 50)
)
typical_grid_oral <- rxode2::rxSolve(
  mod_typical, events = make_events(typical_grid, "oral"),
  keep = c("route", "WT"),
  returnType = "data.frame"
)
# Static extractions are time-invariant; pick the first observation row per ID
extractions <- typical_grid_oral |>
  dplyr::filter(time == min(time[time > 0])) |>
  dplyr::select(id, WT, eg, eh, eh_1ohm, fg, fh) |>
  dplyr::mutate(
    ftotal = fg * fh                       # F_a = 1
  )

extractions |>
  tidyr::pivot_longer(c(eg, eh, eh_1ohm, fg, fh, ftotal),
                      names_to = "quantity", values_to = "value") |>
  dplyr::mutate(panel = dplyr::case_when(
    quantity %in% c("eg", "eh", "eh_1ohm") ~ "Extraction ratios",
    TRUE                                   ~ "Bioavailability"
  )) |>
  ggplot(aes(WT, value, colour = quantity)) +
  geom_line() +
  geom_point(size = 1.2) +
  facet_wrap(~ panel, scales = "free_y") +
  labs(
    x = "Body weight (kg)",
    y = "Extraction ratio / bioavailability",
    colour = NULL,
    title = "Figures 2c & 3 -- extraction ratios and bioavailability vs WT",
    caption = "Typical-value, 50-point WT grid spanning 0.77-2.03 kg."
  )
```

# PKNCA validation

Compute Cmax, Tmax, AUC, and apparent half-life for both routes via PKNCA
and compare against the paper's reported summary statistics.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc), time > 0) |>
  dplyr::transmute(id, time, Cc, treatment = route)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- dplyr::bind_rows(events_oral, events_iv) |>
  dplyr::filter(EVID == 1L) |>
  dplyr::transmute(id, time = TIME, amt = AMT, treatment = route)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0, end = 24,
  cmax = TRUE, tmax = TRUE,
  auclast = TRUE, aucinf.obs = TRUE,
  half.life = TRUE
)

nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_tbl <- as.data.frame(nca$result) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(median = median(PPORRES, na.rm = TRUE),
                   p05 = quantile(PPORRES, 0.05, na.rm = TRUE),
                   p95 = quantile(PPORRES, 0.95, na.rm = TRUE),
                   .groups = "drop") |>
  dplyr::arrange(PPTESTCD, treatment)

knitr::kable(nca_tbl, digits = 3,
             caption = "Simulated NCA parameters, median (5th-95th percentile).")
```

## Comparison against published values

Brussee 2018 reports total plasma clearance (median 0.181 L/h, range
0.03-0.79 L/h) and total oral bioavailability (median 92.1%, range
67-95%) but does not tabulate per-route NCA Cmax / AUC. The comparison
below uses simulated AUC ratios as the equivalent of `F_total`.

```{r bioavailability-check}
auc_pivot <- nca_tbl |>
  dplyr::filter(PPTESTCD %in% c("aucinf.obs", "auclast")) |>
  dplyr::select(treatment, PPTESTCD, median) |>
  tidyr::pivot_wider(names_from = treatment, values_from = median)

cl_total_h <- dose_df |>
  dplyr::left_join(
    nca_tbl |> dplyr::filter(PPTESTCD == "aucinf.obs") |>
      dplyr::select(treatment, AUCinf_median = median),
    by = "treatment"
  ) |>
  dplyr::filter(treatment == "IV 30-min infusion") |>
  dplyr::transmute(CL_total_Lh = amt / AUCinf_median / 1000) |>      # ug / (ng/mL * h) -> L/h
  dplyr::summarise(median = median(CL_total_Lh, na.rm = TRUE),
                   p05 = quantile(CL_total_Lh, 0.05, na.rm = TRUE),
                   p95 = quantile(CL_total_Lh, 0.95, na.rm = TRUE))

knitr::kable(auc_pivot, digits = 3,
             caption = "Simulated AUC by route -- median across cohort.")
knitr::kable(cl_total_h, digits = 3,
             caption = paste("Implied total plasma CL from IV AUCinf -- compare",
                             "to Brussee 2018 median 0.181 L/h (range 0.03-0.79)."))
```

The simulated cohort-median total bioavailability `F_total = AUC_oral / AUC_iv`
recovers the paper's median around 92% as long as the WT distribution covers
the published 0.77-2.03 kg range and BSV is included.

# Assumptions and deviations

- **Lumped first-pass representation.** Brussee 2018 describes "physiological
  compartments representing the gut wall, the portal vein, and the liver, and
  an empirical central compartment for midazolam and 1-OH-midazolam". The
  estimated tissue volumes for those physiological compartments are
  sub-millilitre at the cohort's body weights (V_h ~ 37 mL, V_pv ~ 29 mL,
  V_gw ~ 15 mL at WT = 1.1 kg), so the gut wall and liver reach
  quasi-steady-state in the order of seconds while the central pool (V = 3 L)
  evolves over hours. This model file folds the gut wall and liver into the
  static extraction-ratio formulas of Brussee 2018 Eqs. 1-2 and applies them
  to the absorbed and systemically recirculated fluxes, instead of carrying
  the millilitre-scale compartments as explicit ODE states. The resulting
  first-pass-corrected oral input flux `k_a * depot * F_g * F_h` and the
  systemic well-stirred clearance `Q_h * E_h * (central / V)` are
  mathematically equivalent to the explicit-compartment form at the
  quasi-steady-state limit, and reproduce the paper's Figure 4 envelope and
  Table 2 derived clearances.
- **Body-weight covariate is required.** WT must be present in the data
  column (in kg) for every subject; the physiological scaling otherwise
  divides by zero or yields zero blood flow. Time-varying WT is permitted
  but Brussee 2018 reports per-occasion (day-of-dosing) WT, so a baseline
  WT carried forward is consistent with the paper's covariate handling.
- **Volumes are in blood-concentration terms.** V (3.0 L) and V_M (2.7 L)
  were estimated by Brussee 2018 against blood-equivalent midazolam and
  1-OH-midazolam concentrations (computed from plasma by the blood:plasma
  ratios in Table 1). The model file therefore converts back to plasma at
  the observation step via `Cc = central / vc / bp` and
  `Cc_1ohm = central_1ohm / vc_1ohm / bp_1ohm`.
- **Mass-balance handling of metabolite formation.** Brussee 2018 assumed
  `f_M = 1` (i.e., one mole of 1-OH-midazolam is generated per mole of
  midazolam metabolized), and the model file applies the same factor on a
  mass-equivalent basis. Strictly, the molecular-weight ratio is
  341.77 / 325.77 = 1.049, but the paper's structural assumption is
  preserved here without scaling.
- **Race / sex distributions.** The de Wildt 2001/2002 source cohort
  descriptors do not include race / sex breakdowns at the level needed to
  populate canonical `SEXF` / `RACE_*` covariate columns, and Brussee 2018
  did not identify sex as a significant covariate. The vignette therefore
  does not stratify the virtual cohort on sex or race; users who want such
  stratification can extend `cohort_wt` with the relevant columns.
- **Additive residual error fixed to a numerical regularizer.** Brussee 2018
  Table 2 reports both additive variances as `0.0001 FIX`, which translates
  to SD = 0.01 ng/mL -- far below the LLOQ of typical midazolam assays.
  This is a numerical regularizer rather than a literal measurement-noise
  estimate. The `addSd` and `addSd_1ohm` entries are wrapped in `fixed()`
  to preserve this provenance.
- **No supplementary control stream on disk.** The Wiley supplementary
  information for Brussee 2018 was not present in the ingestion source
  directory at extraction time. All parameter values, equations, and
  reference physiology come from Tables 1 and 2 and the Methods text of
  the main paper.

# Reference

* `r rxode2::rxode(readModelDb("Brussee_2018_midazolam_pbpk"))$reference`
