---
title: "Lidocaine 5% medicated plaster (Bursi 2017)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Lidocaine 5% medicated plaster (Bursi 2017)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Bursi_2017_lidocaine"))$reference`
* Article: <https://doi.org/10.1007/s13318-017-0400-7>
* PubMed Central (open access): <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5597703/>

Bursi et al. (2017) describe a four-compartment parent-metabolite population PK
model for lidocaine and three of its metabolites -- monoethylglycinexylidide
(MEGX), glycinexylidide (GX) and 2,6-xylidine -- measured in serum after
repeated application of the lidocaine 5% medicated plaster in patients with
post-herpetic neuralgia. It is the first population analysis of long-term
exposure to this product.

Two features make the model unusual for a popPK extraction:

1. **Lidocaine has no direct elimination pathway.** The Fig. 1 legend enumerates
   exactly five transfers (k12, k14, k23, k30, k40). There is no k10 and no k20,
   so lidocaine leaves the central compartment only by metabolism, and MEGX
   leaves only by conversion to GX. Lidocaine's apparent elimination rate
   constant is therefore just `k12 + k14 = 0.03 + 0.007 = 0.037 /h`.
2. **Everything about lidocaine is apparent (`/F`).** The plaster input is the
   nominal delivered amount, not the absorbed amount, so `V1 = 1320 L` is a
   `V/F`. The Discussion back-calculates a topical bioavailability of roughly
   5% by comparing the apparent 26 L/kg against a literature IV volume of
   1.3 L/kg.

```{r load-model}
mod <- readModelDb("Bursi_2017_lidocaine")
ui <- rxode2::rxode(mod)

# Four ODE states, four declared endpoints. rxode2 assigns the endpoint
# compartment slots AFTER the ODE states, so observation records in the event
# tables below must carry `cmt = "Cc"` (an endpoint name), not `cmt =
# "central"`. This is the multi-endpoint case: with two or more `~` residual
# lines the endpoint appears in `predDf` and is part of the model definition,
# so nothing is auto-injected and no compartment is renumbered.
ui$state
ui$predDf[, c("var", "cmt", "dvid")]
```

# Population

The analysis pooled 212 patients with post-herpetic neuralgia from two phase III
trials: a randomised-withdrawal trial contributing up to 10 weeks of treatment
and an open-label long-term trial contributing up to 12 months, for up to 14.5
months of follow-up overall (the Discussion notes some subjects reached 18
months). Post-herpetic neuralgia was defined as neuropathic pain persisting at
least 3 months after healing of a herpes zoster rash, with average pain
intensity of at least 4 on the 11-point numeric rating scale at screening; both
trials enrolled patients aged 50 years and older.

Table 1 gives the pooled demographics: median age 72 years (45-92), median
weight 72.5 kg (38-114), median height 165 cm (142-189), median BMI
26.3 kg/m^2 (13.5-44.6), and 119 of 212 patients (56.1%) female. Race,
ethnicity and region are not reported. Sampling was sparse -- up to five
occasions per patient in each trial (Sect. 2.5) -- contributing 1989
concentrations in total, which Table 2 splits as 513 lidocaine, 474 MEGX, 480
GX and 522 2,6-xylidine. The lower limit of quantification was approximately
0.500 ng/mL for all four analytes.

Patients applied up to three plasters simultaneously but not overlapping, for at
most 12 h in each 24-h period. Each plaster contains 700 mg lidocaine and is
assumed to deliver it at a constant 1800 ug/h (Sect. 2.7.7 assumption 1), i.e.
21600 ug per plaster per 12-h application, or about 3% of the plaster's content.

The same information is available programmatically:

```{r population}
str(readModelDb("Bursi_2017_lidocaine")()$population)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Bursi_2017_lidocaine.R`. Collected here for
review:

| Equation / parameter | Value | Source location |
|---|---|---|
| Four-compartment ADVAN5 topology, one compartment per chemical entity | n/a | Sect. 3 para. 2; Fig. 1 and its legend |
| Zero-order plaster input, 1800 ug/h per plaster for 12 h | n/a | Sect. 2.7.7 assumption 1; Sect. 2.4 |
| `hi = hTV * exp(gi)` (log-normal IIV) | n/a | Sect. 2.7.1 eq. (1) |
| `Cij = Chat_ij + e_aij` (additive residual) | n/a | Sect. 2.7.1 eq. (2) |
| `hTV = h1 * COV + h2 * (1 - COV)` (binary covariate switch) | n/a | Sect. 2.7.4 eq. (3) |
| `lk_megx_form` (k12) | fixed 0.03 /h | Table 3, row `k 12 (h-1)` |
| `lk_xyl_form` (k14) | fixed 0.007 /h | Table 3, row `k 14 (h-1)` |
| `lk_gx_form` (k23) | 1.93 /h | Table 3, row `k 23 (h-1)` |
| `lkel_gx_dlvlle2` (k30, <= 2 plasters) | 1.44 /h | Table 3, row `k 30 (h-1) for DLVL B 2` |
| `lkel_gx_dlvlgt2` (k30, 3 plasters) | 2.07 /h | Table 3, row `k 30 (h-1) for DLVL [ 2` |
| `e_tbili_kel_gx` | -0.526 /h | Table 3, row `Effect of BIL [0.53 on k 30` |
| `e_crcl_kel_gx` | -0.32 /h | Table 3, row `Effect of CL CR B 52.7 on k 30` |
| `e_s1a2_kel_gx` | 0.852 /h | Table 3, row `Effect of CYP1A2 substrate on k 30` |
| `e_bmi_kel_gx` | 0.938 /h | Table 3, row `Effect of BMI [27.9 on k 30` |
| `e_alt_kel_gx` | -0.492 /h | Table 3, row `Effect of ALT [11 on k 30` |
| `lkel_xyl_ldhle195` (k40, LDH <= 195) | 0.667 /h | Table 3, row `k 40 (h-1) for LDH B 195` |
| `lkel_xyl_ldhgt195` (k40, LDH > 195) | 0.410 /h | Table 3, row `k 40 (h-1) for LDH [ 195` |
| `e_alt_kel_xyl` | 0.229 /h | Table 3, row `Effect of ALT [11 on k 40` |
| `lvc_dlvlle2` (V1, <= 2 plasters) | 1320 L | Table 3, row `V 1 (L) for DLVL B 2` |
| `lvc_dlvlgt2` (V1, 3 plasters) | 1810 L | Table 3, row `V 1 (L) for DLVL [ 2` |
| `lvc_megx`, `lvc_gx`, `lvc_xyl` | fixed 100 L | Table 3, row `V 2 , V 3 , V 4 (L)`; rationale in Sect. 3 para. 3 |
| `etalkel_gx` | var 0.39 (CV 62.4%) | Table 3, IIV `Proportional on k 30` |
| `etalkel_xyl` | var 0.20 (CV 44.7%) | Table 3, IIV `Proportional on k 40` |
| `etalvc` | var 0.312 (CV 55.9%) | Table 3, IIV `Proportional on V 1` |
| `addSd` | sqrt(364) = 19.1 ug/L | Table 3, `Additive for Lidocaine` |
| `addSd_megx` | sqrt(53.3) = 7.3 ug/L | Table 3, `Additive for MEGX` |
| `addSd_gx` | sqrt(47.9) = 6.9 ug/L | Table 3, `Additive for GX` |
| `addSd_xyl` | sqrt(6.39) = 2.5 ug/L | Table 3, `Additive for 2,6-xylidine` |

Two reading decisions in that table are argued in **Assumptions and deviations**
below: that Table 3's IIV and residual `Estimate` column holds **variances**
(not SDs), and that the `Effect of ...` rows are **additive on the linear 1/h
scale** rather than multiplicative.

# Relationship to the DDMORE sibling model

`nlmixr2lib` already ships `NA_NA_lidocaine`, extracted from DDMORE Foundation
Model Repository entry **DDMODEL00000281**. That bundle's header states that no
linked publication could be identified, and the model was filed under a
placeholder `NA_NA` name with maintainer-chosen default units and no demographics.

**That bundle is this analysis.** The evidence is listed below; the two models
are now cross-linked via `replicate_of` in both files.

| Concordance | Bursi 2017 | DDMODEL00000281 |
|---|---|---|
| Structure | ADVAN5, 4 compartments, transfers k12/k14/k23/k30/k40, no k10/k20 | identical |
| k12, k14 | fixed 0.03, 0.007 | fixed 0.03, 0.007 |
| k23 | 1.93 | 1.93 |
| k30 by dose level | 1.44 / 2.07 | 1.44 / 2.07 |
| k40 by LDH | 0.667 / 0.410 | 0.667 / 0.410 |
| V1 by dose level | 1320 / 1810 | 1320 / 1810 |
| V2, V3, V4 | fixed 100 | fixed 100 |
| Covariate thresholds | DLVL>2, BIL>0.53, CLCR<=52.7, BMI>27.9, ALT>11, LDH>195, CYP1A2 | same, BMI printed as 27.93 |
| Residual variances | 364, 53.3, 47.9, 6.39 | 364, 53.3, 47.9, 6.39 |
| Observation count | 1989 (Table 2 sum) | 1989 |
| Dose record | 1800 ug/h for 12 h per plaster | AMT 21600 / RATE 1800 |
| Institution | co-author J. Grevel, BAST Inc Ltd | licence registered to BAST Inc. Ltd |
| Date | published online 12/01/2017 | run dated 29/11/2016 |

The bundle's run249 is a **neighbouring run**, not the published final model:
five estimates differ in the third significant figure (THETA 6 -0.529 vs -0.526,
THETA 7 -0.319 vs -0.32, THETA 8 0.853 vs 0.852, THETA 9 0.939 vs 0.938, and the
three omegas 0.391/0.200/0.311 vs 0.39/0.2/0.312), and the `.res` listing reports
**325 individuals** against this paper's **212 patients** over the same 1989
observations. The subject-count difference is not explained by anything on disk;
it is recorded here as an open discrepancy rather than resolved.

Use **this** model for the published final estimates and the real units,
population and covariate semantics; use `NA_NA_lidocaine` when you specifically
want the deposited DDMORE artifact reproduced as-deposited.

# Structural checks

These are closed-form identities that follow from the published numbers, so they
are checked at the typical value (`zeroRe()`) with tight tolerances: both sides
use the same parameters and any disagreement is transcription error, not
simulation noise.

## Apparent clearance and apparent volume per kilogram

The Discussion computes `CL/F = Kel * V1` with `Kel = k12 + k14`, obtaining
48.8 L/h for two or fewer plasters and 67.0 L/h for three, and describes the
apparent volume as "about 19 and 26 L/kg (assuming a typical subject of 70 kg)".

```{r structural-clearance}
iniv <- function(p) {
  v <- ui$iniDf$est[ui$iniDf$name == p]
  stopifnot(length(v) == 1L)
  v
}
kel_lidocaine <- exp(iniv("lk_megx_form")) + exp(iniv("lk_xyl_form"))
v1_low <- exp(iniv("lvc_dlvlle2"))
v1_high <- exp(iniv("lvc_dlvlgt2"))

structural <- tibble::tibble(
  Quantity = c(
    "Kel = k12 + k14 (1/h)",
    "CL/F, <= 2 plasters (L/h)", "CL/F, 3 plasters (L/h)",
    "V1/F per 70 kg, <= 2 plasters (L/kg)", "V1/F per 70 kg, 3 plasters (L/kg)"
  ),
  Model = c(kel_lidocaine, kel_lidocaine * v1_low, kel_lidocaine * v1_high,
            v1_low / 70, v1_high / 70),
  Published = c(0.037, 48.8, 67.0, 19, 26),
  Source = c("Sect. 4 (printed as 0.0037; see Errata)", "Sect. 4", "Sect. 4",
             "Sect. 3 ('about 19')", "Sect. 3 ('about 26')")
)
knitr::kable(structural, digits = 4,
             caption = "Apparent clearance and volume against the paper's own arithmetic.")

# The two clearances are printed to three significant figures, so a 0.1%
# tolerance is appropriate. The L/kg values are printed as whole numbers
# ("about 19", "26"), so they get an absolute half-unit tolerance.
stopifnot(
  abs(kel_lidocaine * v1_low - 48.8) / 48.8 < 0.001,
  abs(kel_lidocaine * v1_high - 67.0) / 67.0 < 0.001,
  abs(v1_low / 70 - 19) < 0.5,
  abs(v1_high / 70 - 26) < 0.5
)
```

That both clearances reproduce to within 0.1% is the arbiter for the `k12`
discrepancy discussed in the Errata: only `k12 = 0.03` (Table 3) gives them.

## All rate constants stay positive over every covariate combination

The five additive modifiers on `k30` include three negative values, so a
covariate combination could in principle drive a rate constant to zero or below.
The source NONMEM stream for the sibling DDMORE run carries an
`IF(...LE.0)...=0.0001` floor at each additive step; this model deliberately has
no floor, because none is reachable. The claim is asserted by **enumeration**
over all 2^5 GX combinations and all 2^2 2,6-xylidine combinations rather than
argued in prose.

```{r structural-positivity}
grid_gx <- expand.grid(
  TBILI_HIGH = 0:1, CRCL_LOW = 0:1, S1A2_IND = 0:1,
  BMI_HIGH = 0:1, ALT_HIGH = 0:1, DLVL_HIGH = 0:1
)
grid_gx$k30 <-
  ifelse(grid_gx$DLVL_HIGH == 1, exp(iniv("lkel_gx_dlvlgt2")), exp(iniv("lkel_gx_dlvlle2"))) +
  iniv("e_tbili_kel_gx") * grid_gx$TBILI_HIGH +
  iniv("e_crcl_kel_gx") * grid_gx$CRCL_LOW +
  iniv("e_s1a2_kel_gx") * grid_gx$S1A2_IND +
  iniv("e_bmi_kel_gx") * grid_gx$BMI_HIGH +
  iniv("e_alt_kel_gx") * grid_gx$ALT_HIGH

grid_xyl <- expand.grid(ALT_HIGH = 0:1, LDH_HIGH = 0:1)
grid_xyl$k40 <-
  ifelse(grid_xyl$LDH_HIGH == 1, exp(iniv("lkel_xyl_ldhgt195")), exp(iniv("lkel_xyl_ldhle195"))) +
  iniv("e_alt_kel_xyl") * grid_xyl$ALT_HIGH

cat(sprintf("k30 over %d combinations: min %.4f, max %.4f /h\n",
            nrow(grid_gx), min(grid_gx$k30), max(grid_gx$k30)))
cat(sprintf("k40 over %d combinations: min %.4f, max %.4f /h\n",
            nrow(grid_xyl), min(grid_xyl$k40), max(grid_xyl$k40)))

# Enumerating gate: every reachable typical value must be strictly positive.
# This can go red -- flipping any published sign, or mistyping a magnitude,
# drives the minimum negative.
stopifnot(
  nrow(grid_gx) == 64L, nrow(grid_xyl) == 4L,
  all(grid_gx$k30 > 0), all(grid_xyl$k40 > 0)
)
```

## Metabolite-to-parent AUC ratios are fixed by the rate constants

Because every transfer is first-order and each metabolite has a single inflow,
integrating the ODEs to exhaustion gives exact ratios that depend only on the
rate constants and volumes:

* `AUC(Cc_megx) / AUC(Cc) = (k12 / k23) * (V1 / V2)`
* `AUC(Cc_gx)   / AUC(Cc) = (k12 / k30) * (V1 / V3)` (all MEGX becomes GX)
* `AUC(Cc_xyl)  / AUC(Cc) = (k14 / k40) * (V1 / V4)`

These hold exactly for a typical subject, so they are gated tightly. They are a
strong check on the ODE wiring: mis-routing any transfer breaks at least one.

```{r structural-auc-ratios}
cov_ref <- list(TBILI = 0.2, CRCL = 80, S1A2 = 0, BMI = 25, ALT = 8, LDH = 150)

solve_typical <- function(n_plaster, times, addl = 0L, covs = cov_ref) {
  ev <- rxode2::et(
    amt = 1800 * n_plaster * 12, rate = 1800 * n_plaster,
    cmt = "central", ii = 24, addl = addl
  )
  ev <- rxode2::et(ev, times, cmt = "Cc")
  d <- as.data.frame(ev)
  d$DLVL <- n_plaster
  for (nm in names(covs)) d[[nm]] <- covs[[nm]]
  rxode2::rxSolve(rxode2::zeroRe(mod), d, returnType = "data.frame",
                  useLinCmt = FALSE)
}

# Integrate a single application out far enough for all four analytes to
# return to baseline (lidocaine t1/2 is ln(2)/0.037 = 18.7 h, so 1500 h is
# ~80 half-lives).
tail_sim <- solve_typical(1, seq(0, 1500, by = 0.5))
trap <- function(x, y) sum(diff(x) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)

auc <- vapply(
  c("Cc", "Cc_megx", "Cc_gx", "Cc_xyl"),
  function(v) trap(tail_sim$time, tail_sim[[v]]),
  numeric(1)
)
k12 <- exp(iniv("lk_megx_form")); k14 <- exp(iniv("lk_xyl_form"))
k23 <- exp(iniv("lk_gx_form"))
k30 <- exp(iniv("lkel_gx_dlvlle2")); k40 <- exp(iniv("lkel_xyl_ldhle195"))
vm <- exp(iniv("lvc_megx"))

ratio_tbl <- tibble::tibble(
  Analyte = c("MEGX", "GX", "2,6-xylidine"),
  Simulated = c(auc[["Cc_megx"]], auc[["Cc_gx"]], auc[["Cc_xyl"]]) / auc[["Cc"]],
  `Closed form` = c((k12 / k23) * (v1_low / vm),
                    (k12 / k30) * (v1_low / vm),
                    (k14 / k40) * (v1_low / vm))
) |>
  dplyr::mutate(`% diff` = 100 * (Simulated - `Closed form`) / `Closed form`)
knitr::kable(ratio_tbl, digits = 4,
             caption = "Metabolite:parent AUC ratios, simulated vs closed form.")

# Same parameters on both sides, so the residual is pure trapezoidal /
# truncation error: a tight bound is correct here.
stopifnot(max(abs(ratio_tbl$`% diff`)) < 0.5)
```

## Lidocaine exposure is structurally independent of the metabolite covariates

Five of the seven covariates (bilirubin, creatinine clearance, CYP1A2 substrate,
BMI, ALT) act only on `k30`, and LDH acts only on `k40`. Neither rate constant
appears in the lidocaine ODE, so lidocaine exposure must be **exactly** invariant
to all six. The paper reaches the same conclusion empirically from Fig. 5 ("the
influence of subject factors is very modest ... the change in exposure is very
small"); in the model it is not merely small but identically zero.

```{r structural-invariance}
cov_extreme <- list(TBILI = 5, CRCL = 30, S1A2 = 1, BMI = 35, ALT = 60, LDH = 400)
base_sim <- solve_typical(3, seq(0, 96, by = 0.5), addl = 3L)
pert_sim <- solve_typical(3, seq(0, 96, by = 0.5), addl = 3L, covs = cov_extreme)

rel_lido <- max(abs(base_sim$Cc - pert_sim$Cc)) / max(base_sim$Cc)
rel_gx <- max(abs(base_sim$Cc_gx - pert_sim$Cc_gx)) / max(base_sim$Cc_gx)
cat(sprintf("relative change in lidocaine Cc: %.3e\n", rel_lido))
cat(sprintf("relative change in GX (control):  %.3e\n", rel_gx))

# The difference is not bit-identical: rxode2 integrates all four states with a
# shared adaptive step size, so changing k30 and k40 perturbs the step sequence
# and hence the last digits of `central` too. That is solver noise, not a
# covariate effect -- observed relative change 1.8e-07, versus 1.8e-01 for GX
# under the identical perturbation, six orders of magnitude apart. The bound is
# set at 1e-5 so it sits well above the solver noise (robust to platform and
# tolerance settings) while still failing by four orders of magnitude if any of
# these covariates ever reached the lidocaine compartment.
stopifnot(
  rel_lido < 1e-5,
  # Control arm, so the test above cannot pass vacuously: the same covariate
  # change must visibly move GX.
  rel_gx > 0.01
)
```

# Virtual cohort

Original observed data are not public. The cohort below mirrors the paper's own
simulation design in Sect. 2.7.6 and Fig. 4: one plaster versus three plasters
simultaneously, each applied for 12 h in a 24-h dosing interval, followed over
the first four days of treatment. Covariates are set to the reference
(covariate-unaffected) condition, matching Fig. 5's "population without affecting
covariates"; by the structural check above this choice cannot influence lidocaine
exposure at all.

The paper simulated 500 individuals per arm. This vignette uses 200 per arm, the
package's cap for vignette cohorts; the comparison below is against medians and
robust quantiles, for which 200 is ample.

```{r cohort}
# set.seed() seeds R's RNG, not rxode2's simulation stream, and rxode2
# partitions its streams per solver thread -- so CI draws a different cohort
# than a workstation does and no seed makes them agree. Every assertion on a
# cohort statistic below is written to hold for any cohort this model can
# produce.
set.seed(20170112)
n_per_arm <- 200L

make_arm <- function(n, n_plaster, label, id_offset = 0L) {
  ev <- rxode2::et(
    amt = 1800 * n_plaster * 12, rate = 1800 * n_plaster,
    cmt = "central", ii = 24, addl = 3L
  )
  ev <- rxode2::et(ev, seq(0, 96, by = 0.5), cmt = "Cc")
  ev <- rxode2::et(ev, id = seq_len(n))
  # Materialise before adding covariate columns: assigning into an rxEt object
  # silently drops the column.
  d <- as.data.frame(ev)
  d$id <- d$id + id_offset
  d$DLVL <- n_plaster
  d$TBILI <- cov_ref$TBILI
  d$CRCL <- cov_ref$CRCL
  d$S1A2 <- cov_ref$S1A2
  d$BMI <- cov_ref$BMI
  d$ALT <- cov_ref$ALT
  d$LDH <- cov_ref$LDH
  d$treatment <- label
  d
}

events <- dplyr::bind_rows(
  make_arm(n_per_arm, 1L, "One plaster", id_offset = 0L),
  make_arm(n_per_arm, 3L, "Three plasters", id_offset = 1000L)
)
stopifnot(
  !anyDuplicated(unique(events[, c("id", "time", "evid")])),
  dplyr::n_distinct(events$id) == 2L * n_per_arm
)
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod, events = events, keep = c("treatment", "DLVL"),
  useLinCmt = FALSE
) |>
  as.data.frame()

stopifnot(nrow(sim) > 0L, !all(is.na(sim$Cc)))
```

# Replicate published figures

## Figure 4 -- Cmax after the first and fourth doses

Replicates Fig. 4 of Bursi 2017, which compares simulated lidocaine Cmax after
the first and the fourth dose for (a) one plaster and (b) three plasters.

```{r figure-4, fig.height = 4}
cmax_by_dose <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(dose_no = dplyr::case_when(
    time <= 24 ~ "First dose",
    time >= 72 ~ "Fourth dose",
    TRUE ~ NA_character_
  )) |>
  dplyr::filter(!is.na(dose_no)) |>
  dplyr::group_by(treatment, dose_no, id) |>
  dplyr::summarise(cmax = max(Cc), .groups = "drop") |>
  dplyr::mutate(dose_no = factor(dose_no, c("First dose", "Fourth dose")))

ggplot(cmax_by_dose, aes(dose_no, cmax)) +
  geom_boxplot(outlier.alpha = 0.3) +
  facet_wrap(~treatment) +
  labs(x = NULL, y = "Lidocaine Cmax (ug/L)",
       title = "Figure 4 -- simulated Cmax, first vs fourth dose",
       caption = "Replicates Figure 4 of Bursi 2017.")
```

## Figure 6 -- no accumulation beyond day 4

Fig. 6 shows simulated exposure to lidocaine and its three metabolites over a
year of three-plaster treatment. The Results state that "steady-state conditions
are reached for lidocaine and its metabolites on the fourth day of treatment and
that no accumulation is predicted afterwards". This is checked at the typical
value across three sampling windows: day 1, day 4 and day 365.

```{r figure-6}
long_times <- c(seq(0, 96, by = 0.5), seq(8736, 8760, by = 0.5))
long_sim <- solve_typical(3, long_times, addl = 364L)

window_peak <- function(df, lo, hi) {
  w <- df[df$time >= lo & df$time <= hi, ]
  c(Cc = max(w$Cc), MEGX = max(w$Cc_megx), GX = max(w$Cc_gx), XYL = max(w$Cc_xyl))
}
peaks <- rbind(
  `Day 1` = window_peak(long_sim, 0, 24),
  `Day 4` = window_peak(long_sim, 72, 96),
  `Day 365` = window_peak(long_sim, 8736, 8760)
)
knitr::kable(peaks, digits = 3,
             caption = "Typical-value peak concentration (ug/L) by treatment day, three plasters.")

accum <- peaks["Day 365", ] / peaks["Day 4", ]
cat("Day 365 / Day 4 peak ratio:\n"); print(round(accum, 4))

stopifnot(
  # Day 4 is already at steady state to within 5% for every analyte: this is
  # the paper's claim, restated as a bound. With Kel = 0.037 /h and tau = 24 h
  # the theoretical shortfall at dose 4 is 1 - exp(-4 * 0.037 * 24) = 2.9%,
  # so 5% leaves headroom without being unfalsifiable -- a genuine accumulating
  # analyte would run well past it.
  all(accum < 1.05), all(accum >= 1),
  # Day 1 must be clearly BELOW day 4, or the "steady state by day 4" claim
  # would be vacuous (nothing accumulated in the first place).
  peaks["Day 4", "Cc"] / peaks["Day 1", "Cc"] > 1.3
)

long_sim |>
  dplyr::filter(time <= 96) |>
  dplyr::select(time, Lidocaine = Cc, MEGX = Cc_megx, GX = Cc_gx, `2,6-xylidine` = Cc_xyl) |>
  tidyr::pivot_longer(-time, names_to = "Analyte", values_to = "conc") |>
  ggplot(aes(time, conc, colour = Analyte)) +
  geom_line() +
  labs(x = "Time (h)", y = "Concentration (ug/L)",
       title = "Figure 6 -- first 4 days, three plasters (typical subject)",
       caption = "Replicates the early-time panel of Figure 6 of Bursi 2017.")
```

# PKNCA validation

NCA is run per analyte over the first (0-24 h) and fourth (72-96 h) dosing
intervals, grouped by treatment arm so the results line up with Table 5.

```{r pknca}
nca_one <- function(conc_col, label) {
  sim_nca <- sim |>
    dplyr::filter(!is.na(.data[[conc_col]])) |>
    dplyr::select(id, time, treatment, conc = dplyr::all_of(conc_col))
  # Guarantee a time-zero anchor per subject; pre-dose concentration is 0 for
  # every analyte here. Filtering on `time > 0` or `conc > 0` would drop it and
  # trigger PKNCA's "AUC range starting before the first measurement" warning.
  sim_nca <- dplyr::bind_rows(
    sim_nca,
    sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, conc = 0)
  ) |>
    dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
    dplyr::arrange(id, treatment, time)
  stopifnot(nrow(sim_nca) > 0L)

  dose_df <- events |>
    dplyr::filter(evid == 1) |>
    dplyr::distinct(id, treatment, amt) |>
    tidyr::expand_grid(time = c(0, 24, 48, 72)) |>
    dplyr::select(id, treatment, time, amt)

  conc_obj <- PKNCA::PKNCAconc(sim_nca, conc ~ time | treatment + id)
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

  intervals <- data.frame(
    start = c(0, 72), end = c(24, 96),
    cmax = TRUE, tmax = TRUE, auclast = TRUE
  )
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
  as.data.frame(res) |> dplyr::mutate(analyte = label)
}

nca_all <- dplyr::bind_rows(
  nca_one("Cc", "Lidocaine"),
  nca_one("Cc_megx", "MEGX"),
  nca_one("Cc_gx", "GX"),
  nca_one("Cc_xyl", "2,6-xylidine")
) |>
  dplyr::mutate(
    dose_no = ifelse(start == 0, "First dose", "Fourth dose"),
    group = paste0(treatment, ", ", tolower(dose_no))
  )

nca_summary <- nca_all |>
  dplyr::group_by(analyte, treatment, dose_no, PPTESTCD) |>
  dplyr::summarise(median = stats::median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median)

nca_summary |>
  dplyr::rename(
    "Analyte" = analyte, "Arm" = treatment, "Dose" = dose_no,
    "Cmax (ug/L)" = cmax, "Tmax (h)" = tmax, "AUClast (ug*h/L)" = auclast
  ) |>
  knitr::kable(digits = 2,
               caption = "Median simulated NCA by analyte, arm and dosing interval (n = 200 per arm).")
```

## Comparison against the published Cmax table

Table 5 of Bursi 2017 reports simulated lidocaine Cmax after the first and
fourth doses for one and three plasters. Those are the paper's own model-based
simulation output (500 individuals), so they are a direct comparator for this
model rather than an independent observation.

```{r nca_compare}
published <- tibble::tribble(
  ~group,                          ~cmax,
  "One plaster, first dose",       13.17,
  "One plaster, fourth dose",      21.74,
  "Three plasters, first dose",    29.17,
  "Three plasters, fourth dose",   48.14
)

sim_lido <- nca_all |>
  dplyr::filter(analyte == "Lidocaine", PPTESTCD == "cmax") |>
  dplyr::select(group, cmax = PPORRES)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_lido,
  reference = published,
  by = "group",
  units = c(cmax = "ug/L"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Simulated vs published (Table 5) lidocaine Cmax medians. * differs by >20%.")
```

`ncaComparisonTable()` returns formatted character, so the numeric gates below
are computed separately rather than parsed back out of the table.

```{r nca_gates}
cohort_med <- sim_lido |>
  dplyr::group_by(group) |>
  dplyr::summarise(sim_median = stats::median(cmax), .groups = "drop") |>
  dplyr::inner_join(published, by = "group") |>
  dplyr::mutate(pct_diff = 100 * (sim_median - cmax) / cmax)
stopifnot(nrow(cohort_med) == 4L)  # guard against a silent label mismatch
print(as.data.frame(cohort_med), digits = 4)

# Typical-value (zeroRe) Cmax is deterministic -- no cohort draw enters it --
# so it is gated tightly against Table 5's medians, which for a log-normally
# distributed V1 are the typical values.
typ_cmax <- vapply(c(1, 3), function(np) {
  s <- solve_typical(np, seq(0, 96, by = 0.25), addl = 3L)
  c(first = max(s$Cc[s$time <= 24]), fourth = max(s$Cc[s$time >= 72]))
}, numeric(2))
typ_flat <- c(typ_cmax[, 1], typ_cmax[, 2])
pub_flat <- c(13.17, 21.74, 29.17, 48.14)
typ_pct <- 100 * (typ_flat - pub_flat) / pub_flat
cat("Typical-value Cmax % difference from Table 5 medians:\n")
print(round(typ_pct, 2))

stopifnot(
  # Deterministic: observed range of |% diff| is 0.3-0.9%, bound set at 5% so
  # a mis-transcribed volume, rate constant or dosing rate blows it instantly.
  max(abs(typ_pct)) < 5,
  # Cohort medians carry Monte Carlo noise from a 200-subject draw, so the
  # bound is looser. Observed |% diff| ~ 1-6%.
  max(abs(cohort_med$pct_diff)) < 20,
  # Dose-proportionality claim (Abstract, Sect. 3): Cmax rises LESS than
  # proportionally with plaster count. The ratio of the three-plaster to the
  # one-plaster typical Cmax must sit strictly between 1 and 3.
  typ_cmax["first", 2] / typ_cmax["first", 1] > 1,
  typ_cmax["first", 2] / typ_cmax["first", 1] < 3,
  typ_cmax["fourth", 2] / typ_cmax["fourth", 1] < 3
)
cat(sprintf("Three-plaster : one-plaster Cmax ratio = %.3f (dose 1), %.3f (dose 4); dose ratio is 3.\n",
            typ_cmax["first", 2] / typ_cmax["first", 1],
            typ_cmax["fourth", 2] / typ_cmax["fourth", 1]))
```

## Simulated exposure against the toxicity threshold

The Discussion states that simulated maximum concentrations "remain well below
the level associated with toxicity ([>] 6000 lg/L)". Table 5's own extreme value
across all four scenarios is 445.2 ug/L.

```{r toxicity}
cohort_q <- cmax_by_dose |>
  dplyr::group_by(treatment, dose_no) |>
  dplyr::summarise(p50 = stats::median(cmax), p95 = stats::quantile(cmax, 0.95),
                   max = max(cmax), .groups = "drop")
knitr::kable(cohort_q, digits = 2,
             caption = "Cohort lidocaine Cmax distribution (ug/L) by arm and dosing interval.")

stopifnot(
  # The paper's safety claim.
  max(cohort_q$max) < 6000,
  # A gate that can actually go red: the 95th percentile of the worst arm must
  # stay near Table 5's own 90% CI upper bound for that scenario (124.36 ug/L).
  # Doubling it leaves room for a 200-subject draw while still failing on any
  # order-of-magnitude transcription error.
  max(cohort_q$p95) < 2 * 124.36
)
```

# Assumptions and deviations

* **Table 3's `Estimate` column holds variances, not standard deviations, for
  both the IIV and the residual-error blocks.** Sect. 2.7.1 says the CV% "was
  approximated by the square root of the variance estimate", and the arithmetic
  confirms it in both blocks: `sqrt(0.39) = 0.624` against the printed CV 62.4%,
  `sqrt(0.2) = 0.447` against 44.7%, `sqrt(0.312) = 0.559` against 55.9%; and
  `sqrt(364) = 19.08`, `sqrt(53.3) = 7.30`, `sqrt(47.9) = 6.92`,
  `sqrt(6.39) = 2.53` against the printed SDs 19.1, 7.3, 6.9 and 2.5 ug/L. The
  exact `sqrt()` of each variance is encoded rather than the rounded SD column.

* **The `Effect of ...` rows are additive on the linear 1/h scale.** Two
  independent tells: Table 3 gives each such row the unit `(h-1)`, where a
  multiplicative factor would be dimensionless; and three of the five GX effects
  are negative, which cannot be a value for a rate constant. They are therefore
  the `(h1 - h2)` increments of the Sect. 2.7.4 eq. (3) switch, with the DLVL
  stratum supplying the baseline. The DLVL and LDH rows, by contrast, print two
  **parallel baselines** and are encoded as stratum switches.

* **`k12` is 0.03 /h, not the 0.003 /h printed in Sect. 4.1.** The paper carries
  three mutually inconsistent statements. Table 3 says `Fixed to 0.03`; Sect. 4
  says "lidocaine K el was estimated at 0.0037 h-1 (sum of k 12 and k 14 )"; and
  Sect. 4.1 Limitations says k12 and k14 were fixed at "0.003 and 0.007 h-1,
  respectively". The apparent clearances the paper computes in the same paragraph
  arbitrate: `48.8 / 1320 = 0.03697` and `67.0 / 1810 = 0.03702`, so the intended
  `Kel` is 0.037 /h, which is `0.03 + 0.007`. Table 3's value is used;
  `0.003 + 0.007 = 0.010` would give a clearance of 13.2 L/h, and the "0.0037"
  is a decimal-point slip for 0.037. The structural-checks section asserts this.

* **The bilirubin threshold's unit is recorded as mg/dL although the Table 3
  footnote says umol/L.** A threshold of 0.53 umol/L is roughly an order of
  magnitude below any measurable total bilirubin (reference range 5-21 umol/L),
  whereas 0.53 mg/dL sits inside the normal range (0.3-1.2 mg/dL). The threshold
  *value* is transcribed verbatim; only the unit label is corrected, and the
  discrepancy is recorded in the model's `covariateData$TBILI$notes`. Supplying
  bilirubin in umol/L to this model would put every subject above the threshold.

* **All of the covariate thresholds are distributional cohort splits, not
  clinical cutoffs.** Sect. 3 describes simulating subjects with "laboratory
  safety parameters from the 30th percentile". ALT > 11 U/L sits below the usual
  adult reference range (about 7-56 U/L) and BMI > 27.9 kg/m^2 near the cohort
  median of 26.3. Users should not read these as hepatic-impairment or obesity
  criteria.

* **`S1A2` is encoded as a plain binary here (0/1), matching Sect. 2.7.4
  ("Concomitant medication was expressed as binary data"), whereas the sibling
  `NA_NA_lidocaine` uses the DDMORE dataset's integer code in which level 3 is
  the "yes" level.** Both are registered against the one canonical; supply a 0/1
  column to this model.

* **Co-medication and all covariates are time-fixed.** Sect. 2.7.7 assumptions 3
  and 5: covariate values were held constant over the entire observation time and
  all co-medications were assumed in use throughout, regardless of start or stop
  dates.

* **No floor is imposed on the covariate-adjusted rate constants.** The DDMORE
  sibling's NONMEM stream carries an `IF(...LE.0)...=0.0001` guard at each
  additive step; it is omitted here because the enumerating check above shows the
  minimum reachable typical value is 0.102 /h for `k30` and 0.410 /h for `k40`.
  A user who re-fits this model with different data should restore the guard.

* **The absorption step is not modelled.** Sect. 2.7.7 assumption 1 assumes each
  plaster delivers lidocaine at a constant 1800 ug/h, so the model has no depot
  and dosing is a zero-order input straight into `central` supplied through the
  event table's `rate` column. Consequently every lidocaine parameter is apparent
  (`V/F`, `CL/F`), and the long apparent half-life of 18.7 h reflects
  absorption-rate limitation (flip-flop kinetics), not lidocaine's true
  disposition half-life of 1.5-2 h after IV administration.

* **The 1800 ug/h delivery rate is per plaster.** The paper states it once, in
  the singular ("plasters were assumed to deliver lidocaine at a constant rate
  (i.e., 1800 lg/h)"), without saying explicitly whether it is per plaster or per
  application. Per plaster is the reading used, and it is confirmed
  quantitatively: it reproduces all four of Table 5's Cmax medians to within 1%,
  whereas a per-application reading would under-predict the three-plaster arm
  threefold. It is also consistent with the DDMORE dataset's `RATE 1800` records
  and with Sect. 1's statement that about 3% of an applied plaster's 700 mg
  reaches the circulation (`1800 * 12 / 700000 = 3.1%`).

* **Cohort size is 200 per arm against the paper's 500.** Comparisons are made on
  medians and robust quantiles; the extremes in Table 5 (Min, Max) are not gated,
  because the extreme of a random cohort is not reproducible across rxode2 builds
  or solver-thread counts.

* **Race, ethnicity and region are not reported by the paper** and are absent from
  the `population` metadata rather than imputed.

* **The subject count disagrees with the DDMORE sibling** (212 patients here
  versus 325 individuals in the DDMODEL00000281 `.res` listing) over the same
  1989 observations. Nothing on disk explains the difference; it is recorded as an
  open discrepancy, not resolved.

* **No erratum or corrigendum was located** for this article on the publisher's
  page or in PubMed Central as of this extraction.

* **Every parameter value comes from the paper's Table 3.** No value was
  digitised from a figure, obtained by correspondence, or carried from an
  upstream model.
