---
title: "Teicoplanin (Butragueno-Laiseca 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Teicoplanin (Butragueno-Laiseca 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-load}
ui <- rxode2::rxode(readModelDb("ButraguenoLaiseca_2025_teicoplanin"))
```

* Citation: `r ui$reference`
* Description: `r ui$description`
* Article: <https://doi.org/10.1093/jac/dkaf012>
* Supplement: `dkaf012_supplementary_data.docx`, available from the Europe PMC
  open-access record for [PMC11879232](https://europepmc.org/article/PMC/PMC11879232)

# Population

Twenty-six critically ill children treated with teicoplanin in a single
paediatric intensive care unit (Hospital General Universitario Gregorio
Maranon, Madrid) were enrolled; twelve of them were undergoing continuous
kidney replacement therapy (CKRT) in continuous venovenous haemodiafiltration
modality. Median (range) weight was 8 kg (4.3-44 kg) and median age 14 months
(3 months to 13 years). Most patients were in the post-operative period of a
congenital cardiopathy, nearly all were mechanically ventilated, and the
median PRISM III score was 6 (no CKRT) and 8 (CKRT). Baseline characteristics
are reported separately for the 15 patients without and the 12 with a
haemofilter in Butragueno-Laiseca 2025 Table 1; the two group sizes sum to 27
rather than 26 because one patient started treatment without CKRT and required
CKRT after the fifth dose, contributing to both groups (Table 1 footnote a).

All patients received the standard paediatric regimen: three loading doses of
10 mg/kg every 12 h, followed 24 h later by a maintenance dose of 10 mg/kg
every 24 h (3.3 mg/kg every 24 h in CKRT patients), each given as a 5-min
intravenous infusion. A total of 173 concentrations were modelled - 72 plasma
samples from non-CKRT patients and 51 pre-filter plus 50 post-filter samples
from CKRT patients - all above the 1.0 mg/L limit of quantification and
ranging from 2.1 to 86.6 mg/L.

The same information is available programmatically via the model's
`population` metadata:

```{r population}
str(ui$population, max.level = 1)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in
`inst/modeldb/specificDrugs/ButraguenoLaiseca_2025_teicoplanin.R`. The table
below collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl_renal` (theta_CLR) | 0.169 L/h | Table 2 (RSE 12%; SIR 95% CI 0.139-0.208) |
| `lcl_hemodialysis` (theta_CLKRT) | 0.119 L/h | Table 2 (RSE 8%; SIR 95% CI 0.103-0.136) |
| `lvc` (theta_V1) | 1.56 L | Table 2 (RSE 17%; SIR 95% CI 1.03-1.98) |
| `lvp` (theta_V2) | 3.03 L | Table 2 (RSE 30%; SIR 95% CI 2.10-4.53) |
| `lq` (theta_CLD) | 0.292 L/h | Table 2 (RSE 53%; SIR 95% CI 0.167-0.614) |
| `e_wt_cl_renal` | 0.75 (fixed) | Results, Covariate selection ("exponents of 1 and 0.75 for V1 and CLR") |
| `e_wt_vc` | 1 (fixed) | Results, Covariate selection |
| reference weight | 8 kg | Table 2 Parameter model column `(WGT/8)`; cohort median weight, Results |
| `e_filt_sa_med_cl_hemodialysis` | 3.58 | Table 2, theta_FILT Med (RSE 13%; SIR 95% CI 2.78-4.40) |
| `e_filt_sa_large_cl_hemodialysis` | 5.04 | Table 2, theta_FILT Large (RSE 14%; SIR 95% CI 3.80-6.26) |
| theta_FILT Small = 1 (reference) | 1 | Table 2 |
| `etalcl_renal` | 30.4% CV | Table 2 IIV column (RSE 32%; eta-shrinkage 33%) |
| `etalcl_hemodialysis` | 14.1% CV | Table 2 IIV column (RSE 19%; eta-shrinkage 45%) |
| `etalvc` | 39.2% CV | Table 2 IIV column (RSE 45%; eta-shrinkage 40%) |
| omega^2 = log(1 + CV^2) | n/a | Table 2 footnote (CV% = sqrt(exp(omega^2) - 1) x 100) |
| `expSd` | 0.301 | Table 2, plasma residual error (RSE 8%; SIR 95% CI 0.27-0.34) |
| `expSd_Cpostfilter` | 0.333 | Table 2, post-filter residual error (RSE 13%; SIR 95% CI 0.28-0.39) |
| Log-additive residual -> `lnorm()` | n/a | Methods, PK analysis ("additive error model in the logarithmic scale") |
| Two-compartment structure | n/a | Results, Base population model; Figure S2; Supplementary material ("one-compartment model performed significantly worse ... three-compartment model did not improve the fit") |
| CLNR removed | n/a | Results, Base population model ("CLNR could not be identified and was therefore removed") |
| CLR = 0 during CKRT | n/a | Table 2 footnote a; Supplementary material, Base population model ("thus, CL equals CLKRT") |
| No IIV on V2 / CLD; no eta covariance | n/a | Results, Base population model |
| `Cpostfilter = Cc * (1 - CLKRT / phi_PlCorr)` | n/a | Supplementary material, Pharmacokinetic analysis (displayed equation) |
| `phi_PlCorr = phi_Blood x BPR` | n/a | Supplementary material, Pharmacokinetic analysis (BPR reading documented in Errata) |
| 5-min IV infusion | n/a | Methods, Dosing sampling and analytical method |

# Structural checks

These checks are deterministic consequences of the encoded parameters and can
be evaluated without simulating: each is a statement the paper makes about its
own model, recomputed from the packaged `ini()` values.

```{r structural-checks}
th <- setNames(ui$theta, names(ui$theta))

theta_clr   <- exp(th[["lcl_renal"]])
theta_clkrt <- exp(th[["lcl_hemodialysis"]])
theta_v1    <- exp(th[["lvc"]])
ref_wt      <- 8

# Discussion: "The model predicts an increase in 0.016 L/h and 0.19 L,
# respectively, for an increment of 1 kg of body weight."
d_clr_dwt <- theta_clr * th[["e_wt_cl_renal"]] / ref_wt   # d/dWT at WT = 8 kg
d_v1_dwt  <- theta_v1  * th[["e_wt_vc"]]       / ref_wt

# Discussion: "the estimate of CLKRT (0.119 L/h) was 30% lower than CLR".
pct_lower <- 100 * (theta_clr - theta_clkrt) / theta_clr

structural <- tibble::tibble(
  Check = c(
    "dCLR/dWT at 8 kg (L/h per kg)",
    "dV1/dWT at 8 kg (L per kg)",
    "CLKRT below CLR (%)",
    "CLKRT, medium 0.6 m2 filter (L/h)",
    "CLKRT, large 1.2 m2 filter (L/h)"
  ),
  Model = c(
    d_clr_dwt, d_v1_dwt, pct_lower,
    theta_clkrt * th[["e_filt_sa_med_cl_hemodialysis"]],
    theta_clkrt * th[["e_filt_sa_large_cl_hemodialysis"]]
  ),
  Published = c(0.016, 0.19, 30, 0.119 * 3.58, 0.119 * 5.04),
  Source = c(
    "Discussion", "Discussion", "Discussion",
    "Table 2 (theta_CLKRT x theta_FILT Med)",
    "Table 2 (theta_CLKRT x theta_FILT Large)"
  )
)

# Each published figure is quoted to two significant figures, so the model
# value must fall inside the rounding interval of the printed value: within
# half a unit of the last published digit. The V1 slope is the tight case --
# 1.56 / 8 is exactly 0.195, which the paper prints as 0.19, so the model
# value sits precisely on the interval boundary and the comparison needs an
# epsilon to survive binary floating point (0.195 - 0.19 evaluates to
# 0.005000000000000004).
matches_published <- function(model, published, half_ulp) {
  abs(model - published) <= half_ulp + 1e-9
}

stopifnot(
  matches_published(d_clr_dwt, 0.016, 0.0005),
  matches_published(d_v1_dwt,  0.19,  0.005),
  matches_published(pct_lower, 30,    1)
)
# The slopes are also exact analytic consequences of the encoded parameters.
stopifnot(
  isTRUE(all.equal(d_clr_dwt, 0.169 * 0.75 / 8)),
  isTRUE(all.equal(d_v1_dwt,  1.56 / 8))
)

structural |>
  dplyr::mutate(Model = signif(Model, 3), Published = signif(Published, 3)) |>
  knitr::kable(caption = "Structural consequences of the encoded parameters versus the values Butragueno-Laiseca 2025 states for its own model.")
```

The post-filter observation equation is checked the same way. With the CKRT
cohort medians of Table 1 (blood flow 50 mL/min, haematocrit 29.3%) the plasma
flow entering the haemofilter is `50 x 0.06 x (1 - 0.293) = 2.12` L/h, so the
single-pass extraction ratio `CLKRT / phi_PlCorr` rises with filter size:

```{r postfilter-check}
plasma_flow <- 50 * 0.06 * (1 - 29.3 / 100)   # mL/min -> L/h, times plasma fraction

tibble::tibble(
  `Filter` = c("Small (0.2 m2)", "Medium (0.6 m2)", "Large (1.2 m2)"),
  `CLKRT (L/h)` = theta_clkrt *
    c(1, th[["e_filt_sa_med_cl_hemodialysis"]], th[["e_filt_sa_large_cl_hemodialysis"]]),
  `Extraction ratio` = theta_clkrt *
    c(1, th[["e_filt_sa_med_cl_hemodialysis"]], th[["e_filt_sa_large_cl_hemodialysis"]]) / plasma_flow
) |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) signif(x, 3))) |>
  dplyr::mutate(`Cpost / Cpre` = signif(1 - `Extraction ratio`, 3)) |>
  knitr::kable(caption = "Single-pass haemofilter extraction implied by the supplementary post-filter equation at the Table 1 median blood flow and haematocrit.")
```

The resulting 6-28% single-pass extraction is consistent with the paper's
observation that post-filter concentrations are only modestly below pre-filter
concentrations, and with the Discussion note that teicoplanin's high protein
binding means "theoretically only 10% of the drug should be eliminated".

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the simulation design of Butragueno-Laiseca 2025 (Methods, Percentage of
target attainment): body weight is the only covariate simulated, sampled
within the three published weight bands (3-10, 11-30, 31-60 kg). Four regimens
are simulated - the standard regimen and the paper's recommended alternative,
for CKRT and non-CKRT patients (Table S1).

**Haemofilter size is held at the reference (small, 0.2 m2) level in these
target-attainment simulations**, which is how the paper ran them: "In
simulations, the only covariate that was simulated was body weight, given that
filter size was considered as a surrogate of body weight. Haemofilter settings
were not needed to simulate plasma concentrations" (Results, Probability of
target attainment). That sentence admits two readings - derive filter size
from the simulated weight, or drop the filter term entirely - and they give
very different answers. The published target-attainment percentages
discriminate between them, and both readings are simulated side by side in
[Which reading of the filter-size sentence?](#which-reading-of-the-filter-size-sentence)
below: holding the filter at its reference level reproduces the published
values and deriving it from weight does not, so that is the design used here.
The filter-size covariate is still encoded in the model - it is what Table 2
estimates - and is exercised separately in its own cohort.

```{r cohort}
set.seed(20250107)

n_per_arm <- 100L

weight_bands <- tibble::tribble(
  ~wt_group,  ~wt_lo, ~wt_hi,
  "3-10 kg",     3,     10,
  "11-30 kg",   11,     30,
  "31-60 kg",   31,     60
)

regimens <- tibble::tribble(
  ~regimen,                              ~ckrt, ~ld_mgkg, ~md_mgkg,
  "No CKRT: 3 x 10 q12h + 10 q24h",          0,       10,     10.0,
  "No CKRT: 3 x 15 q12h + 15 q24h",          0,       15,     15.0,
  "CKRT: 3 x 10 q12h + 3.3 q24h",            1,       10,      3.3,
  "CKRT: 3 x 15 q12h + 10 q24h",             1,       15,     10.0
)

arms <- tidyr::crossing(regimens, weight_bands)

# Observation grid: fine over the first hour so the 5-min infusion peak is
# resolved, then every 30 min out to the 72 h trough.
obs_times <- sort(unique(c(seq(0, 1, by = 0.05), seq(1, 72, by = 0.5))))
dose_times <- c(0, 12, 24, 48)   # 3 loading doses q12h, first maintenance 24 h later
inf_dur    <- 5 / 60             # 5-min infusion (Methods, Dosing)

make_arm <- function(regimen, ckrt, ld_mgkg, md_mgkg,
                     wt_group, wt_lo, wt_hi,
                     n = n_per_arm, id_offset = 0L) {
  subj <- tibble::tibble(
    id              = id_offset + seq_len(n),
    WT              = exp(runif(n, log(wt_lo), log(wt_hi))),
    RRT_CRRT_ACTIVE = ckrt,
    # Reference (small, 0.2 m2) haemofilter throughout -- see the note above.
    FILT_SA_MED     = 0,
    FILT_SA_LARGE   = 0,
    # Blood flow and haematocrit enter only the post-filter observation.
    # Table 1 medians (range) for the CKRT group: 50 (34-150) mL/min and
    # 29.3% (26.3-37.5%). Non-CKRT subjects carry BFR = 0 (no circuit).
    BFR             = if (ckrt == 1) exp(runif(n, log(34), log(150))) else 0,
    HCT             = runif(n, 26.3, 37.5),
    regimen         = regimen,
    wt_group        = wt_group
  )

  doses <- subj |>
    tidyr::crossing(time = dose_times) |>
    dplyr::mutate(
      amt  = WT * ifelse(time < 48, ld_mgkg, md_mgkg),
      rate = amt / inf_dur,
      evid = 1L,
      cmt  = "central",
      dvid = NA_integer_
    )

  # Both model outputs (Cc and Cpostfilter) are algebraic transforms of the
  # single `central` state, so rxode2 assigns them endpoint slots AFTER the
  # ODE states and an observation row carrying cmt = "central" cannot be
  # matched to either endpoint. Observation rows therefore select the endpoint
  # with dvid = 1 (Cc) and leave cmt unset; rxode2 returns BOTH observables as
  # columns on every observation row.
  obs <- subj |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(
      amt  = NA_real_, rate = NA_real_, evid = 0L,
      cmt  = NA_character_, dvid = 1L
    )

  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- do.call(
  dplyr::bind_rows,
  lapply(seq_len(nrow(arms)), function(i) {
    a <- arms[i, ]
    do.call(make_arm, c(as.list(a), list(id_offset = (i - 1L) * n_per_arm)))
  })
)

stopifnot(!anyDuplicated(events[, c("id", "time", "evid")]))
stopifnot(length(unique(events$id)) == nrow(arms) * n_per_arm)

nrow(events)
```

# Simulation

```{r simulate}
mod <- readModelDb("ButraguenoLaiseca_2025_teicoplanin")

sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep   = c("regimen", "wt_group", "WT", "RRT_CRRT_ACTIVE"),
  # rxode2's automatic ODE -> linCmt conversion mis-maps the endpoints of
  # multi-output models; see known-vignette-failure-patterns.md pattern 5b.
  useLinCmt = FALSE
) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc))
```

# Replicate published figures

```{r figure-s1, fig.cap = "Simulated plasma concentration-time profiles by weight band and regimen. Replicates the layout of Figure S1 of Butragueno-Laiseca 2025 (median with 95% prediction interval; therapeutic range 10-60 mg/L in dashed red)."}
sim |>
  dplyr::group_by(regimen, wt_group, time) |>
  dplyr::summarise(
    Q025 = quantile(Cc, 0.025),
    Q50  = quantile(Cc, 0.50),
    Q975 = quantile(Cc, 0.975),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q025, ymax = Q975), alpha = 0.25) +
  geom_line() +
  geom_hline(yintercept = c(10, 60), linetype = "dashed", colour = "red") +
  facet_grid(wt_group ~ regimen) +
  scale_y_log10() +
  labs(
    x = "Time since first dose (h)",
    y = "Teicoplanin plasma concentration (mg/L)",
    title = "Simulated concentration-time profiles",
    caption = "Replicates the layout of Figure S1 of Butragueno-Laiseca 2025."
  ) +
  theme(strip.text = element_text(size = 7))
```

# Haemofilter size and the post-filter observable

The target-attainment cohort above holds the filter at its reference size, so
the filter-size covariate and the post-filter observable are exercised by a
separate typical-value cohort: three subjects differing only in filter size,
all at the CKRT-group median weight of 9.11 kg (Table 1) on the standard CKRT
regimen, with inter-individual variability zeroed so the comparison is exact.

```{r filter-cohort}
filter_arms <- tibble::tribble(
  ~filter,           ~FILT_SA_MED, ~FILT_SA_LARGE,
  "Small (0.2 m2)",             0,              0,
  "Medium (0.6 m2)",            1,              0,
  "Large (1.2 m2)",             0,              1
)

# Typical-value replication: IIV is zeroed so the three arms differ ONLY in
# filter size and the fold-changes can be checked exactly against Table 2
# rather than compared as medians of separately-drawn random samples.
mod_typ <- rxode2::zeroRe(mod)

filt_subj <- filter_arms |>
  dplyr::mutate(
    id              = dplyr::row_number(),
    WT              = 9.11,
    RRT_CRRT_ACTIVE = 1,
    BFR             = 50,
    HCT             = 29.3
  )

filt_events <- dplyr::bind_rows(
  filt_subj |>
    tidyr::crossing(time = dose_times) |>
    dplyr::mutate(
      amt = WT * ifelse(time < 48, 10, 3.3), rate = amt / inf_dur,
      evid = 1L, cmt = "central", dvid = NA_integer_
    ),
  filt_subj |>
    tidyr::crossing(time = seq(0, 72, by = 1)) |>
    dplyr::mutate(
      amt = NA_real_, rate = NA_real_, evid = 0L,
      cmt = NA_character_, dvid = 1L
    )
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(!anyDuplicated(filt_events[, c("id", "time", "evid")]))

sim_filt <- rxode2::rxSolve(
  mod_typ, events = filt_events, keep = "filter", useLinCmt = FALSE
) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc), time > 0)

filt_summary <- sim_filt |>
  dplyr::group_by(filter) |>
  dplyr::summarise(
    cl_krt   = dplyr::first(cl_hemodialysis),
    retained = dplyr::first(Cpostfilter / Cc),
    cmin72   = Cc[time == 72],
    .groups  = "drop"
  ) |>
  dplyr::mutate(filter = factor(filter, levels = filter_arms$filter)) |>
  dplyr::arrange(filter)

# With IIV zeroed these are exact identities, not approximate agreements:
# the fold-changes ARE the Table 2 theta_FILT multipliers, and the retained
# fraction IS the supplementary equation evaluated at the Table 1 medians.
cl_small <- filt_summary$cl_krt[filt_summary$filter == "Small (0.2 m2)"]
stopifnot(
  isTRUE(all.equal(cl_small, 0.119)),
  isTRUE(all.equal(
    filt_summary$cl_krt / cl_small,
    c(1, 3.58, 5.04)
  )),
  isTRUE(all.equal(
    filt_summary$retained,
    1 - filt_summary$cl_krt / plasma_flow
  ))
)

filt_summary |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) signif(x, 3))) |>
  dplyr::rename(
    "Filter"              = filter,
    "CLKRT (L/h)"         = cl_krt,
    "Cpost / Cpre"        = retained,
    "Cmin at 72 h (mg/L)" = cmin72
  ) |>
  knitr::kable(caption = "Typical-value effect of haemofilter surface area at a fixed 9.11 kg body weight on the standard CKRT regimen. Clearance rises exactly 3.58-fold (medium) and 5.04-fold (large) relative to the small reference filter -- the theta_FILT multipliers of Table 2 -- single-pass retention across the filter falls, and the 72 h trough falls with it.")
```

The retention ratios reproduce the deterministic values computed above from
the supplementary equation, and the fold-changes in `CLKRT` reproduce the
`theta_FILT` multipliers of Table 2 exactly; both are asserted in the chunk
above rather than left to visual inspection. Note that the single-pass
extraction and the whole-body clearance are separate quantities: even a large
filter removes only about 28% of the drug per pass, but it does so
continuously, which is what makes `CLKRT` five times the small-filter value.

# PKNCA validation

Butragueno-Laiseca 2025 evaluates exposure over the 48-72 h window, i.e.
between the first and second maintenance doses (Methods, Percentage of target
attainment: "AUC24 represents the area under the plasma concentration-versus-
time curve obtained between 48 and 72 h post initiation of teicoplanin
therapy"). The PKNCA interval below matches that window.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, regimen, wt_group)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | regimen + wt_group + id)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, regimen, wt_group)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | regimen + wt_group + id)

intervals <- data.frame(
  start     = 48,
  end       = 72,
  cmax      = TRUE,
  tmax      = TRUE,
  cmin      = TRUE,
  auclast   = TRUE,
  half.life = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- PKNCA::pk.nca(nca_data)

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(regimen, wt_group, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

nca_wide |>
  dplyr::group_by(regimen, wt_group) |>
  dplyr::summarise(
    cmax      = median(cmax),
    cmin      = median(cmin),
    auclast   = median(auclast),
    half.life = median(half.life, na.rm = TRUE),
    .groups   = "drop"
  ) |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) signif(x, 3))) |>
  dplyr::rename(
    "Regimen"                 = regimen,
    "Weight band"             = wt_group,
    "Cmax 48-72 h (mg/L)"     = cmax,
    "Cmin 48-72 h (mg/L)"     = cmin,
    "AUC48-72 (mg*h/L)"       = auclast,
    "t1/2 (h)"                = half.life
  ) |>
  knitr::kable(caption = "Median PKNCA-derived exposure over the 48-72 h dosing interval, by regimen and weight band.")
```

The terminal half-life recovered over this interval is consistent with the
paper's rationale for once-daily maintenance dosing ("one of the main
advantages of teicoplanin, its relatively long half-life, allowing once-daily
dosing", Discussion). The paper does not tabulate NCA parameters, so no
`ncaComparisonTable()` row-by-row comparison is possible; the published
quantities that *are* numeric are the target-attainment percentages compared
below.

## Comparison against published target attainment

The paper reports the percentage of simulated patients reaching a trough
concentration above 10 mg/L at 72 h under the standard regimen, and the
AUC24/MIC > 800 target at an MIC of 1 mg/L.

```{r pta}
cmin72 <- sim |>
  dplyr::filter(abs(time - 72) < 1e-8) |>
  dplyr::select(id, regimen, wt_group, Cmin72 = Cc)

pta <- nca_wide |>
  dplyr::left_join(cmin72, by = c("id", "regimen", "wt_group")) |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(
    `PTA Cmin > 10 mg/L (%)`      = 100 * mean(Cmin72 > 10),
    `PTA Cmin > 15 mg/L (%)`      = 100 * mean(Cmin72 > 15),
    `PTA AUC24/MIC > 800, MIC 1 (%)` = 100 * mean(auclast > 800),
    `Cmin > 60 mg/L (toxicity, %)`   = 100 * mean(Cmin72 > 60),
    .groups = "drop"
  )

published_pta <- tibble::tribble(
  ~regimen,                            ~`Published PTA Cmin > 10 mg/L (%)`, ~`Published source`,
  "No CKRT: 3 x 10 q12h + 10 q24h",    66,   "Discussion (mean across weight groups)",
  "No CKRT: 3 x 15 q12h + 15 q24h",    NA,   "Figure 3a: > 90% target",
  "CKRT: 3 x 10 q12h + 3.3 q24h",      84,   "Discussion (mean across weight groups)",
  "CKRT: 3 x 15 q12h + 10 q24h",       NA,   "Figure 3b: > 90% target"
)

pta |>
  dplyr::left_join(published_pta, by = "regimen") |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) signif(x, 3))) |>
  dplyr::rename("Regimen" = regimen) |>
  knitr::kable(caption = "Simulated versus published probability of target attainment. Published values for the standard regimens are the cross-weight-group means quoted in the Discussion; for the recommended regimens the paper states only that target attainment exceeds 90% (Figure 3).")
```

Both standard regimens reproduce the paper's headline conclusion and its
numbers: 61% versus a published 66% without a haemofilter, and 77% versus a
published 84% with one, both short of the 90% threshold the authors adopt. The
two recommended regimens clear that threshold (91% and 96%), matching the
paper's proposal of `3 x 15 mg/kg q12h + 15 mg/kg q24h` for children without
CKRT and `3 x 15 mg/kg q12h + 10 mg/kg q24h` for children on CKRT.

The counter-intuitive ordering - CKRT patients attaining the trough target
*more* often than non-CKRT patients despite receiving a third of the
maintenance dose - is a property of the published model, not an artifact:
`CLKRT` carries no allometric term, so it stays at 0.119 L/h regardless of
body size, while `CLR` grows as `WT^0.75`. Heavier children on CKRT therefore
accumulate, and the simulation reproduces the paper's corresponding safety
warning that "to prevent Cmin > 60 mg/L, a loading dose exceeding 15 mg/kg in
patients weighing more than 10 kg undergoing haemofiltration should be
avoided" (Discussion): 20% of simulated CKRT patients exceed 60 mg/L on the
standard regimen and 46% on the recommended one, entirely from the two heavier
weight bands, whereas no non-CKRT regimen approaches that threshold.

The AUC24/MIC > 800 attainment shown here is a simulated quantity and is not
comparable to the 0% and 3.7% the paper reports for its 26 *observed*
patients, who carried their real filter sizes and covariates rather than the
reference-filter simulation design.

Exact agreement with the published percentages is not expected: the paper
simulated 1000 profiles per scenario from its own weight distribution
(Figure S1), which is not published in numeric form, whereas this vignette
samples weight log-uniformly within each published band and uses 100 subjects
per arm. Both standard-regimen percentages land within about 8 points of the
published values, and both recommended regimens reproduce the paper's
conclusion that they clear the 90% threshold for a trough above 10 mg/L.

## Which reading of the filter-size sentence?

The cohort above holds the haemofilter at its reference (small) size. The
alternative reading of the paper's sentence - that filter size was *derived*
from the simulated weight, since the paper assigned filter size by weight band
(small 3-10 kg, medium 10-30 kg, large 30-60 kg; Results, Patient population) -
is simulated here on exactly the same subjects, so the two readings differ only
in the value of `FILT_SA_MED` / `FILT_SA_LARGE`.

```{r filter-discriminator}
filter_by_band <- tibble::tribble(
  ~wt_group,  ~med_derived, ~large_derived,
  "3-10 kg",             0,              0,
  "11-30 kg",            1,              0,
  "31-60 kg",            0,              1
)

# Same CKRT subjects, same weights, same dosing -- only the filter indicators
# change, so this is a paired comparison rather than a re-simulation.
events_derived <- events |>
  dplyr::filter(RRT_CRRT_ACTIVE == 1) |>
  dplyr::left_join(filter_by_band, by = "wt_group") |>
  dplyr::mutate(FILT_SA_MED = med_derived, FILT_SA_LARGE = large_derived) |>
  dplyr::select(-med_derived, -large_derived)

stopifnot(
  nrow(events_derived) == sum(events$RRT_CRRT_ACTIVE == 1),
  !anyNA(events_derived$FILT_SA_MED),
  !anyNA(events_derived$FILT_SA_LARGE)
)

pta_derived <- rxode2::rxSolve(
  mod, events = events_derived,
  keep = c("regimen", "wt_group"), useLinCmt = FALSE
) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc), abs(time - 72) < 1e-8) |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(derived = 100 * mean(Cc > 10), .groups = "drop")

published_disc <- tibble::tribble(
  ~regimen,                        ~published, ~published_label,
  "CKRT: 3 x 10 q12h + 3.3 q24h",          84, "84 (Discussion)",
  "CKRT: 3 x 15 q12h + 10 q24h",           90, "> 90 (Figure 3b)"
)

discriminator <- pta |>
  dplyr::select(regimen, reference = `PTA Cmin > 10 mg/L (%)`) |>
  dplyr::inner_join(pta_derived, by = "regimen") |>
  dplyr::inner_join(published_disc, by = "regimen")

# The design decision this table justifies: for the standard CKRT regimen --
# the one scenario with a published point estimate rather than a ">= 90%"
# statement -- the reference-filter reading must be the closer of the two.
std <- discriminator |>
  dplyr::filter(regimen == "CKRT: 3 x 10 q12h + 3.3 q24h")
stopifnot(
  nrow(std) == 1L,
  abs(std$reference - std$published) < abs(std$derived - std$published)
)

discriminator |>
  dplyr::mutate(dplyr::across(c(reference, derived), \(x) signif(x, 3))) |>
  dplyr::select(-published) |>
  dplyr::rename(
    "Regimen"                              = regimen,
    "Filter held at reference (%)"         = reference,
    "Filter derived from weight band (%)"  = derived,
    "Published (%)"                        = published_label
  ) |>
  knitr::kable(caption = "The two readings of the paper's filter-size sentence, simulated on the same subjects. Holding the filter at its reference size reproduces the published target attainment; deriving it from the weight band does not.")
```

Deriving the filter from weight costs 30-55 percentage points of target
attainment because `CLKRT` carries no allometric term of its own, so a medium or large
filter multiplies clearance 3.6- or 5-fold in exactly the heavier children
whose weight-proportional dose would otherwise have compensated. The residual
gap between the reference-filter simulation and the published 84% is
attributable to the weight distribution: this vignette samples log-uniformly
within each band, while the paper sampled from its own (non-published)
distribution shown in Figure S1.

# Assumptions and deviations

- **Post-filter plasma flow (`BPR`).** The supplementary equation is
  `Cpost = Cpre * (1 - CLKRT / phi_PlCorr)` with `phi_PlCorr = phi_Blood x BPR`,
  where BPR is described only as the "blood to plasma ratio"; the paper does
  not print its numerical definition. It is encoded here as the plasma
  fraction of blood, `1 - HCT/100`, which is the reading that makes the
  Results statement that "CKRT parameters and haematocrit were included in the
  model" true of this equation, and which yields physically plausible
  single-pass extraction ratios of 6-28% across the three filter sizes.
- **Numerical floors in the post-filter equation.** `model()` clamps the
  plasma flow to a minimum of 0.001 L/h and the retained fraction
  `1 - CLKRT/phi_PlCorr` to a minimum of 0.001. Neither floor binds anywhere
  in the covariate ranges the paper reports; they exist so that non-CKRT
  subjects (who have no extracorporeal circuit and therefore `BFR = 0`) do not
  divide by zero, and so that an extreme sampled combination of a large filter
  with a low blood flow cannot produce a negative concentration.
- **Residual-error units.** Table 2 labels the residual-error rows
  "Residual error, log (mg/mL)". A residual standard deviation on the log
  scale is unitless, and every concentration in the paper is reported in
  mg/L; the "(mg/mL)" label is treated as a reporting slip and the values
  0.301 and 0.333 are encoded as log-scale (lnorm) residual SDs.
- **Number of observations.** The Abstract states 172 concentration
  measurements while the Results state 173. The Results figure is the one
  consistent with the reported breakdown (72 plasma + 51 pre-filter + 50
  post-filter = 173) and is the value recorded in the model's `population`
  metadata.
- **Allometric exponents encoded as fixed.** The exponents of 0.75 (CLR) and
  1 (V1) are reported in the Results text as values the authors "incorporated"
  into the allometric expression, and Table 2 gives no RSE or confidence
  interval for either, so both are wrapped in `fixed()`.
- **Pre-filter and plasma concentrations share one prediction.** The paper
  estimates one residual magnitude for the combined "plasma/pre-filter" group
  and a second for post-filter samples, so `Cc` represents both the plasma and
  the pre-filter concentration and `Cpostfilter` is the separate post-filter
  observable.
- **Haemofilter size in the target-attainment simulations.** The paper states
  that "the only covariate that was simulated was body weight, given that
  filter size was considered as a surrogate of body weight", which could mean
  either that filter size was derived from the simulated weight or that the
  filter term was not applied at all. The two readings give target-attainment
  percentages roughly 30 points apart, and only the second reproduces the
  published values (see the Virtual cohort section for the side-by-side
  comparison), so the target-attainment cohort holds the filter at its
  reference (small) level. This affects only the vignette's simulation design:
  the model file encodes the filter-size covariate exactly as Table 2 reports
  it, and a separate cohort exercises it.
- **Cohort covariate distributions.** Body weight is sampled log-uniformly
  within the three published weight bands; the paper's own weight
  distribution (Figure S1) is not published numerically. Blood flow is set to
  the Table 1 median in the filter-size cohort and sampled log-uniformly over
  the Table 1 range
  (34-150 mL/min) and haematocrit uniformly over the CKRT-group range
  (26.3-37.5%); neither affects the plasma predictions, only the post-filter
  observable.
- **Target-attainment simulations use `Cc` (the individual prediction) rather
  than a residual-error-perturbed observation.** The paper does not state
  whether its target-attainment calculation included residual variability.
- **Sex, race and ethnicity are not covariates.** Sex percentages are reported
  per group only (75% male without and 40% male with a haemofilter) and no
  overall value is given; race and ethnicity are not reported.
