---
title: "Manganese monkey and human PBPK (Campbell 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Manganese monkey and human PBPK (Campbell 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Campbell_2023_manganese_monkey_pbpk"))$reference`
* Article: <https://doi.org/10.1093/toxsci/kfac123>
* Supplement (complete ODE listing + Table S1): <https://doi.org/10.1093/toxsci/kfac123> (Supplementary Material)

Campbell et al. recast the Schroeter (2011) monkey and human manganese PBPK models
so that manganese binds to saturable tissue sites **rapidly and reversibly**,
following the structure Yoon et al. (2019) had established for the adult rat.
In the older models the association/dissociation of bound manganese was slow and
was deliberately made the rate-limiting step for loss of tissue manganese. Here
the binding and dissociation rate constants are large (dissociation half-lives of
minutes), a single equilibrium dissociation constant `KD` of roughly 0.5 uM is
shared across tissues, and the tissue-to-tissue differences live entirely in the
maximal binding capacity `Bmax`. Free manganese accumulates only once those sites
saturate, which is what produces the sharply nonlinear tissue dose response above
about 0.1 mg/m3.

The paper's central methodological claim is a **cross-species transfer**: the
tissue influx, efflux and binding-capacity parameters were estimated against
monkey data only, then carried over unchanged (with allometric scaling) into the
human model, where only the dietary absorbed fraction and the biliary clearance
were re-fitted. This library therefore ships two files that share every
chemical-specific parameter and differ only in physiology and gut handling.

```{r load}
monkey <- readModelDb("Campbell_2023_manganese_monkey_pbpk")
human <- readModelDb("Campbell_2023_manganese_human_pbpk")
mkUi <- rxode2::rxode2(monkey)
huUi <- rxode2::rxode2(human)
c(monkey_states = length(mkUi$state), human_states = length(huUi$state))
```

Each species carries 65 ODEs: a 34-state bulk-manganese system and a 31-state
`54Mn` radiotracer twin. The tracer is not a separable linear sub-system --
tracer and bulk manganese **compete for the same binding sites**, because the
free-capacity term is `Bmax - bound_<organ> - bound_<organ>_mn54`.

# Population

```{r population}
mkPop <- monkey()$population
huPop <- human()$population
tibble::tibble(
  Field = c("Species", "Weight", "Dose range"),
  Monkey = c(mkPop$species, mkPop$weight_range, mkPop$dose_range),
  Human = c(huPop$species, huPop$weight_median, huPop$dose_range)
) |>
  knitr::kable()
```

Monkey data are the Dorman (2006a) 90-day inhalation study (4-6 animals per
exposure concentration, MnSO4 aerosol at 0, 0.06, 0.3 and 1.5 mg Mn/m3, 6 h/day
5 days/week) plus four whole-body `54Mn` tracer studies. Human data are almost
entirely whole-body `54Mn` retention studies plus plasma manganese responses to
controlled dietary change; there are **no human tissue time-course data**, so
predicted human tissue concentrations rest on the cross-species parameter
transfer rather than on direct observation.

# Source trace

Every value in both files carries an in-file comment naming its source location.
The table below is the reviewer-facing summary.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Quantity, ~`Source location`,
  "Body weight, tissue volume fractions, brain-region fractions, nasal surface areas and thickness, cardiac output, alveolar ventilation, tissue blood flows", "Table 1 (monkey and human columns)",
  "Brain-region volume fractions for the monkey (2x right-hemisphere weights)", "Supplementary Table S1",
  "Tissue influx / efflux diffusion rate constants (KIN*, KOUT*)", "Table 2, 'Cellular transport'; optimised to monkey, reused for human",
  "Association rate constant (0.182 per ug/L per h, all 8 tissues)", "Table 2, 'Tissue binding constants'; Yoon et al. 2019",
  "Equilibrium dissociation constants (20 or 25 ug/L)", "Table 2 (given as ka x KD); Discussion confirms '0.37 or 0.46 uM (20 or 25 ug/l)'",
  "Maximal binding capacities (BMAX*)", "Table 2, 'Maximal storage tissue capacities'",
  "Regional deposition fractions and epithelial clearances", "Table 2, 'Fraction deposited and olfactory bulb translocation' (airflow split taken from Methods -- see Errata)",
  "GI parameters (FDIETUP, KGI, FENT, KENT, KFECES)", "Table 2, 'GI parameters' (human KGI corrected -- see Errata)",
  "Biliary clearance and its Hill-type induction (KBILEC, KBINDUC, KM, SLOPE)", "Table 2 (continued); human KBILEC also confirmed in the Figure 9 caption",
  "Full ODE structure, tissue vascular fraction (3%), output definitions", "Supplementary Material model listing",
  "Exposure protocol (6 h/day, 5 days/week, 90 days)", "Figures 3 and 6 captions",
  "Plasma = 20% of blood manganese", "Figure 10 caption"
) |>
  knitr::kable()
```

## Units

```{r units-table, echo = FALSE}
tibble::tribble(
  ~Quantity, ~Unit, ~Note,
  "Time", "h", "the supplement integrates in hours; figures plot days",
  "State amounts", "ug Mn", "every ODE state is an amount, not a concentration",
  "Volumes", "L", "body weight in kg is taken as L (density 1)",
  "Tissue concentration output", "ug/g", "amount / (volume in L x 1000); the units of Figures 3 and 6",
  "Blood concentration (Cc)", "ug/L", "arterial whole blood",
  "Dietary intake", "mg/day", "converted internally to ug/h by x1000/24",
  "Inhaled air concentration", "mg/m3", "numerically identical to ug/L of air",
  "Association rate constant", "per ug/L per h", "so ka x KD has units 1/h",
  "Biliary induction KM", "ug/g", "Table 2 labels it ug/l; the supplement compares it against ug/g -- see Errata"
) |>
  knitr::kable()
```

# Dimensional analysis

Two checks worth stating explicitly, because both are places a transcription
error would hide.

**Binding.** `ka` is `per ug/L per h` and `KD` is `ug/L`, so the dissociation
rate `kd = ka * KD` is `1/h`. The bound-pool ODE is
`-kd * bound + ka * (free sites, ug) * (free conc, ug/L)`, which is
`ug/h` on both terms. The equilibrium bound amount is therefore
`Bmax * C / (KD + C)` -- an ordinary saturable isotherm.

```{r dimensional-binding}
ka <- 0.182
kdVals <- c(globus_pallidus = 20, cerebellum = 20, other = 20, pituitary = 20,
            liver = 25, lung = 25, bone = 25, olfactory_bulb = 25)
mwMn <- 54.938  # g/mol
tibble::tibble(
  tissue = names(kdVals),
  KD_ug_L = kdVals,
  KD_uM = round(kdVals / mwMn, 3),
  kd_per_h = round(ka * kdVals, 3),
  kd_half_life_min = round(log(2) / (ka * kdVals) * 60, 1)
) |>
  knitr::kable()
```

The `KD` values in uM are 0.364 and 0.455 against the Discussion's printed "0.37
or 0.46 uM", and both are near the "approximately 0.5 uM" the paper attributes to
Yoon et al. (2019). The 25 ug/L value reproduces its printed 0.46 exactly; the
20 ug/L value computes to 0.364, whose rounding interval does not quite reach the
printed 0.37, so the paper has rounded 0.364 up by one in the last digit. The
assertion below therefore checks the *conversion* rather than the paper's
rounding -- it still goes red on a wrong molecular weight or a factor-of-ten unit
slip, which is what it is there to catch.

```{r dimensional-binding-check}
kdUM <- unname(kdVals[c("globus_pallidus", "liver")]) / mwMn
stopifnot(
  # The ug/L -> uM conversion must land on the Discussion's printed pair to
  # within its last printed digit (0.364 vs 0.37; 0.455 vs 0.46).
  all(abs(kdUM - c(0.37, 0.46)) < 0.01),
  # ... and every tissue must sit near the paper's "approximately 0.5 uM".
  all(kdVals / mwMn > 0.3), all(kdVals / mwMn < 0.6)
)
```

The Discussion additionally describes these dissociation rate constants as
"equivalent to half-lives of 4-5 min". From the printed `ka` and `KD` the
half-lives are 9.1-11.4 min, so that particular sentence does not follow from
the printed parameters; it is a Discussion characterisation and no model
parameter depends on it (recorded in the Errata).

**Pituitary.** `bmax_pituitary` is 0.55 ug/L of tissue, essentially zero next to
the other tissues' 100-7900 ug/L. That is not a typo: it makes pituitary
manganese almost entirely *free*, so at steady state the pituitary tracks blood
with a fixed amplification set only by the influx/efflux ratio and the
volume bookkeeping. Typing that closed form from the printed Table 1 and Table 2
values gives an independent check on the ODE implementation.

```{r dimensional-pituitary}
# Closed form from PRINTED values only (not read back from the model object):
#   at steady state kin*A_brainblood = kout*A_pit, and brain blood is 3% of brain
#   volume, so C_pit / C_brainblood = (kin/kout) * 0.03 / f_pituitary.
amp_printed <- (0.016 / 0.012) * 0.03 / 0.00037
amp_printed
```

That predicts a pituitary:blood ratio of about 108, so a baseline blood
manganese of 0.010 ug/g implies roughly 1.08 ug/g in the pituitary -- which is
what Figure 3H shows at day 0. The same arithmetic reproduces the Figure 3H peak
(blood 0.075 ug/g x 108 = 8.1 ug/g of free plus vascular manganese). This is
checked numerically against the solved model below.

# Monkey: steady state on the background diet

Both files start every state at zero, exactly as the supplement's listing does,
with the note that "state variables must be initialized to the steady state
amount or the model must be run to steady state before assessing any additional
exposure". So the first step of any simulation is a burn-in.

Campbell 2023 does not report the monkey dietary manganese intake numerically
(the supplement's `INFAC` intake factor, needed for the ppm route, is not given).
We therefore anchor the monkey scenario on the paper's own reported baseline:
the dietary rate that places the model at the Figure 3F day-0 blood manganese of
0.010 ug/g. This is a **calibrated scenario input, not a published parameter**,
and it is the only quantity in this vignette that is fitted rather than read.

```{r monkey-burnin}
solveMn <- function(ui, params = list(), times, inits = NULL) {
  m <- ui
  if (length(params)) m <- do.call(rxode2::ini, c(list(m), params))
  args <- list(m, events = rxode2::et(times), atol = 1e-9, rtol = 1e-9,
               maxsteps = 1e6)
  if (!is.null(inits)) args$inits <- inits
  as.data.frame(do.call(rxode2::rxSolve, args))
}
steadyState <- function(ui, params = list(), days = 1500) {
  r <- solveMn(ui, params, seq(0, days * 24, by = 24))
  unlist(r[nrow(r), rxode2::rxState(ui)])
}
# Solve for the dietary rate reproducing the reported baseline blood Mn.
bloodTarget <- 0.010  # ug/g, Figure 3F at day 0
dietObj <- function(logDiet) {
  r <- solveMn(mkUi, list(diet_mn_mgd = exp(logDiet)), seq(0, 1500 * 24, by = 24))
  tail(r$Cc, 1) / 1000 - bloodTarget
}
mkDiet <- exp(stats::uniroot(dietObj, c(log(1e-3), log(200)), tol = 1e-9)$root)
mkInit <- steadyState(mkUi, list(diet_mn_mgd = mkDiet))
round(mkDiet, 2)
```

```{r monkey-ss-checks}
mkSs <- solveMn(mkUi, list(diet_mn_mgd = mkDiet), seq(0, 1500 * 24, by = 24))
mkLast <- mkSs[nrow(mkSs), ]
# 1. Steady state: no drift over the final 100 days of the burn-in.
tailRows <- mkSs$time >= (1400 * 24)
drift <- diff(range(mkSs$Ctot_globus_pallidus[tailRows])) /
  mean(mkSs$Ctot_globus_pallidus[tailRows])
# 2. Mass balance: everything absorbed is either in the body or in faeces.
# 3. The printed-value pituitary amplification, checked against the solve.
ampSolved <- mkLast$Ctot_pituitary / (mkLast$Cc / 1000)
c(relative_drift = signif(drift, 3),
  mass_balance = signif(mkLast$Mn_balance, 8),
  amp_solved = signif(ampSolved, 4),
  amp_printed = signif(amp_printed, 4))
```

```{r monkey-ss-assert}
stopifnot(
  # Steady state reached: less than 0.1% spread over the last 100 days.
  drift < 1e-3,
  # Mass balance is an identity of the ODE system, so it should be exact.
  abs(mkLast$Mn_balance - 1) < 1e-6,
  # The solved pituitary:blood amplification must match the closed form typed
  # from the printed Table 1 / Table 2 values. The solve includes the vascular
  # term the closed form omits, so allow 15%.
  abs(ampSolved / amp_printed - 1) < 0.15
)
```

## Baseline tissue concentrations against Figure 3 day 0

```{r monkey-baseline}
fig3Baseline <- tibble::tribble(
  ~output,                 ~tissue,             ~digitised,
  "Ctot_liver",            "Liver",             2.40,
  "Ctot_lung",             "Lung",              0.17,
  "Ctot_bone",             "Bone",              0.13,
  "Ctot_other",            "Rest of body",      0.16,
  "Ctot_globus_pallidus",  "Globus pallidus",   0.55,
  "Ctot_pituitary",        "Pituitary",         1.00,
  "Ctot_cerebellum",       "Cerebellum",        0.43,
  "Ctot_olfactory_bulb",   "Olfactory bulb",    0.33
)
baseCmp <- fig3Baseline |>
  mutate(model = as.numeric(mkLast[output]),
         pct_diff = 100 * (model - digitised) / digitised)
baseCmp |>
  select(-output) |>
  rename("Tissue" = tissue, "Figure 3 day 0 (ug/g)" = digitised,
         "Model (ug/g)" = model, "% difference" = pct_diff) |>
  knitr::kable(digits = 3)
```

```{r monkey-baseline-assert}
stopifnot(
  # One calibrated scalar (the diet) has to carry eight tissues at once, so
  # this is a structural check on the binding/transport parameters, not a fit.
  abs(median(baseCmp$pct_diff)) < 15,
  stats::quantile(abs(baseCmp$pct_diff), 0.9) < 35
)
```

# Monkey: 90-day inhalation, replicating Figure 3

```{r monkey-inhalation}
mkInh <- solveMn(
  mkUi,
  list(diet_mn_mgd = mkDiet, conc_air_mgm3 = 1.5, expo_duration_days = 90),
  seq(0, 180 * 24, by = 2), inits = mkInit
)
# The published curves are read at the end of the last exposure period.
lastExposure <- max(which(mkInh$expo_on == 1))
mkPeak <- mkInh[lastExposure, ]
mkInh$day <- mkInh$time / 24
mkInh |>
  select(day, Liver = Ctot_liver, `Globus pallidus` = Ctot_globus_pallidus,
         Pituitary = Ctot_pituitary, Cerebellum = Ctot_cerebellum,
         `Olfactory bulb` = Ctot_olfactory_bulb, Bone = Ctot_bone) |>
  pivot_longer(-day, names_to = "Tissue", values_to = "conc") |>
  ggplot(aes(day, conc)) +
  geom_line(linewidth = 0.3) +
  facet_wrap(~Tissue, scales = "free_y") +
  labs(x = "Time (days)", y = "Total Mn (ug/g)",
       title = "Replicates Figure 3 of Campbell 2023 (monkey, 1.5 mg Mn/m3)",
       subtitle = "90-day inhalation, 6 h/day, 5 days/week, then 90 days of clearance") +
  theme_bw()
```

The sawtooth is the 6 h/day exposure cycle; the rapid post-exposure decline in
the brain regions is the behaviour the rapid-dissociation structure was
introduced to capture.

The strongest available check is against the paper's **own published model
curves** at the end of the last exposure -- this compares our implementation
against theirs, so a transcription or structural error shows up directly.

```{r monkey-fig3-peak}
fig3Curve <- tibble::tribble(
  ~output,                 ~tissue,            ~digitised,
  "Ctot_liver",            "Liver",             4.30,
  "Ctot_lung",             "Lung",              0.78,
  "Ctot_bone",             "Bone",              0.30,
  "Ctot_other",            "Rest of body",      0.47,
  "Ctot_globus_pallidus",  "Globus pallidus",   3.10,
  "Ctot_pituitary",        "Pituitary",         6.20,
  "Ctot_cerebellum",       "Cerebellum",        1.15,
  "Ctot_olfactory_bulb",   "Olfactory bulb",    3.00
)
peakCmp <- fig3Curve |>
  mutate(model = as.numeric(mkPeak[output]),
         pct_diff = 100 * (model - digitised) / digitised)
peakCmp |>
  select(-output) |>
  rename("Tissue" = tissue, "Figure 3 model curve (ug/g)" = digitised,
         "This implementation (ug/g)" = model, "% difference" = pct_diff) |>
  knitr::kable(digits = 3)
```

```{r monkey-fig3-assert}
stopifnot(
  # Zero free parameters here: the diet was fixed by the baseline above and
  # the exposure is fully specified, so this is our ODEs vs the authors' ODEs.
  abs(median(peakCmp$pct_diff)) < 10,
  max(abs(peakCmp$pct_diff)) < 25,
  # Mass balance must still hold under a cycling inhalation exposure.
  abs(mkPeak$Mn_balance - 1) < 1e-6
)
```

# Monkey: dose response and binding-site saturation, replicating Figure 6

```{r monkey-doseresponse}
doseLevels <- c(0, 0.06, 0.3, 1.5)
drRows <- lapply(doseLevels, function(cAir) {
  r <- solveMn(mkUi,
               list(diet_mn_mgd = mkDiet, conc_air_mgm3 = cAir,
                    expo_duration_days = 90),
               seq(0, 90 * 24, by = 2), inits = mkInit)
  i <- if (cAir == 0) nrow(r) else max(which(r$expo_on == 1))
  data.frame(conc_air = cAir,
             globus_pallidus = r$Ctot_globus_pallidus[i],
             pituitary = r$Ctot_pituitary[i],
             liver = r$Ctot_liver[i],
             blood = r$Cc[i] / 1000,
             pct_sites_free_gp = r$Pfree_globus_pallidus[i],
             pct_sites_free_liver = r$Pfree_liver[i])
})
dr <- dplyr::bind_rows(drRows)
dr |>
  rename("Air Mn (mg/m3)" = conc_air, "Globus pallidus (ug/g)" = globus_pallidus,
         "Pituitary (ug/g)" = pituitary, "Liver (ug/g)" = liver,
         "Blood (ug/g)" = blood, "% GP sites free" = pct_sites_free_gp,
         "% liver sites free" = pct_sites_free_liver) |>
  knitr::kable(digits = 4)
```

```{r monkey-doseresponse-assert}
stopifnot(
  # Monotone increase in the target region with exposure.
  all(diff(dr$globus_pallidus) > 0),
  # Globus-pallidus binding sites are progressively consumed.
  all(diff(dr$pct_sites_free_gp) < 0),
  # The globus pallidus runs essentially saturated: under 5% of its capacity is
  # free even on the background diet alone, and under 1% at 1.5 mg/m3.
  dr$pct_sites_free_gp[dr$conc_air == 0] < 5,
  dr$pct_sites_free_gp[dr$conc_air == 1.5] < 1,
  # The liver, by contrast, keeps a large reserve at every exposure level, so
  # it buffers rather than accumulating free Mn.
  all(dr$pct_sites_free_liver > 40),
  all(dr$pct_sites_free_liver > 8 * dr$pct_sites_free_gp),
  # Supralinearity: the 0.3 -> 1.5 fold change must exceed the 0.06 -> 0.3 one,
  # even though both are 5-fold increments in air concentration.
  (dr$globus_pallidus[4] / dr$globus_pallidus[3]) >
    (dr$globus_pallidus[3] / dr$globus_pallidus[2]),
  # And the target region must rise far more steeply than the liver over the
  # same exposure range -- the contrast Figure 6 draws between panels A and G.
  (dr$globus_pallidus[4] / dr$globus_pallidus[2]) >
    2 * (dr$liver[4] / dr$liver[2])
)
```

That block is the paper's mechanism stated as a test, and the numbers sharpen it
in a way worth noting: the globus pallidus is **already about 96% saturated on
the background diet alone** (4.1% of capacity free), so it has almost no spare
buffering reserve and any additional manganese delivered to it appears as *free*
manganese. The liver keeps roughly 65% of its much larger capacity free at
baseline and still 44% at 1.5 mg/m3. That asymmetry is why globus-pallidus
manganese rises 4.3-fold across the exposure range while liver manganese rises
only 1.5-fold -- exactly the contrast Figure 6 draws between panels A and G, and
the reason the brain is the target of concern.

# Human: steady state and plasma manganese

```{r human-burnin}
huInit <- steadyState(huUi, days = 4000)
huSs <- solveMn(huUi, times = seq(0, 4000 * 24, by = 24))
huLast <- huSs[nrow(huSs), ]
c(plasma_ug_L = signif(huLast$Cplasma, 4),
  blood_ug_L = signif(huLast$Cc, 4),
  mass_balance = signif(huLast$Mn_balance, 8))
```

```{r human-plasma-assert}
stopifnot(
  # Figure 10's observed average plasma Mn across six dietary periods spans
  # 0.72-1.25 ug/L; a 3 mg/day background diet must land inside that.
  huLast$Cplasma > 0.70, huLast$Cplasma < 1.30,
  abs(huLast$Mn_balance - 1) < 1e-6
)
```

## Replicating Figure 10: stepped dietary manganese

Freeland-Graves (1994) fed five men six successive controlled diets. This is the
one place a scenario input has to vary with time, which is done by supplying
`diet_mn_mgd` as a data column -- rxode2 honours a data column that shadows an
`ini()` parameter.

```{r human-fig10}
dietSchedule <- tibble::tribble(
  ~from_day, ~to_day, ~diet_mn_mgd,
  0,   21,  2.36,
  21,  42,  2.89,
  42,  63,  2.06,
  63,  101, 1.21,
  101, 112, 3.79,
  112, 125, 2.65
)
dietSchedule |>
  rename("From (day)" = from_day, "To (day)" = to_day,
         "Diet (mg Mn/day)" = diet_mn_mgd) |>
  knitr::kable()

fig10Times <- seq(0, 125 * 24, by = 6)
dietAt <- function(tHours) {
  d <- tHours / 24
  idx <- findInterval(d, dietSchedule$from_day, rightmost.closed = TRUE)
  idx[idx < 1] <- 1
  idx[idx > nrow(dietSchedule)] <- nrow(dietSchedule)
  dietSchedule$diet_mn_mgd[idx]
}
fig10Ev <- data.frame(id = 1L, time = fig10Times, evid = 0L, amt = 0,
                      diet_mn_mgd = dietAt(fig10Times))
# Start from the steady state of the first dietary period, not of 3 mg/day.
init10 <- steadyState(huUi, list(diet_mn_mgd = dietSchedule$diet_mn_mgd[1]),
                      days = 4000)
fig10 <- as.data.frame(rxode2::rxSolve(
  huUi, fig10Ev, inits = init10, atol = 1e-9, rtol = 1e-9, maxsteps = 1e6
))
fig10$day <- fig10$time / 24
# The caption reports one average plasma concentration per dietary period, not
# per sampling day, so each observed mean is placed at its period MIDPOINT.
obs10 <- dietSchedule |>
  mutate(day = (from_day + to_day) / 2,
         plasma = c(1.05, 1.25, 0.95, 0.72, 1.10, 0.97),
         sd = c(0.15, 0.30, 0.55, 0.15, 0.20, 0.25))
ggplot(fig10, aes(day, Cplasma)) +
  geom_line() +
  geom_pointrange(data = obs10,
                  aes(day, plasma, ymin = plasma - sd, ymax = plasma + sd),
                  inherit.aes = FALSE, size = 0.3) +
  labs(x = "Time (days)", y = "Plasma Mn (ug/L)",
       title = "Replicates Figure 10 of Campbell 2023",
       subtitle = "Simulated plasma Mn (20% of blood) against Freeland-Graves 1994 period means") +
  theme_bw()
```

```{r human-fig10-assert}
fig10Cmp <- obs10 |>
  mutate(model = approx(fig10$day, fig10$Cplasma, day)$y,
         ratio = model / plasma,
         within_sd = abs(model - plasma) <= sd)
fig10Cmp |>
  select(diet_mn_mgd, day, plasma, sd, model, ratio, within_sd) |>
  rename("Diet (mg/day)" = diet_mn_mgd, "Period midpoint (day)" = day,
         "Observed plasma (ug/L)" = plasma, "SD" = sd, "Model (ug/L)" = model,
         "Model/observed" = ratio, "Within 1 SD" = within_sd) |>
  knitr::kable(digits = 3)
c(median_fold_error = signif(1 / median(fig10Cmp$ratio), 4))
stopifnot(
  # The whole simulated trajectory must stay in the physiological band for
  # plasma Mn.
  all(fig10$Cplasma > 0.4), all(fig10$Cplasma < 2.0),
  # The discriminating check: the model must pick out the SAME dietary period
  # as the observed data for both the lowest and the highest plasma Mn. This
  # would fail on a sign error, a mis-ordered schedule, or a diet term that
  # did not actually drive blood Mn.
  which.min(fig10Cmp$model) == which.min(fig10Cmp$plasma),
  which.max(fig10Cmp$model) == which.max(fig10Cmp$plasma),
  # Level agreement. The model runs systematically low here (see Errata), so
  # this bounds the offset rather than asserting it away.
  1 / median(fig10Cmp$ratio) < 1.6,
  max(pmax(fig10Cmp$ratio, 1 / fig10Cmp$ratio)) < 2.0
)
```

The model reproduces *which* dietary period gives the lowest and the highest
plasma manganese, and the shape of the response, but sits about 1.37-fold below
the observed period means throughout. The largest single discrepancy is the
3.79 mg/day period, where the observed mean has risen to 1.10 ug/L by the
period midpoint while the model has only reached 0.62: the model's body pool
equilibrates more slowly than the data suggest over an 11-day window. Both
points are recorded in the Errata; nothing was tuned to reduce them.

# Human: whole-body 54Mn retention

The tracer is dosed as events. An intravenous dose is a bolus into
`venous_mn54`; an oral dose **splits**, with `f_diet_uptake` of the dose going
directly to `liver_mn54` (the absorbed fraction) and the remainder to
`gut_lumen_mn54`, mirroring the supplement's partition of `KDIETR`.

```{r human-tracer-fn}
tracerRun <- function(route, fDietUp = 0.06, clBile = 0.051, days = 65) {
  pars <- list(f_diet_uptake = fDietUp, cl_bile = clBile, dose_mn54 = 1)
  ui <- do.call(rxode2::ini, c(list(huUi), pars))
  init <- steadyState(huUi, pars, days = 4000)
  ev <- if (route == "iv") {
    rxode2::et(amt = 1, cmt = "venous_mn54", time = 0)
  } else {
    rbind(rxode2::et(amt = fDietUp, cmt = "liver_mn54", time = 0),
          rxode2::et(amt = 1 - fDietUp, cmt = "gut_lumen_mn54", time = 0))
  }
  as.data.frame(rxode2::rxSolve(
    ui, rxode2::et(ev, seq(0, days * 24, by = 6)), inits = init,
    atol = 1e-10, rtol = 1e-10, maxsteps = 1e7
  ))
}
retentionAt <- function(r, days) approx(r$time / 24, r$Retention_mn54, days)$y
```

## Intravenous tracer, replicating Figure 7

```{r human-fig7}
ivRun <- tracerRun("iv")
ivRun$day <- ivRun$time / 24
fig7 <- tibble::tibble(day = c(10, 20, 30, 40, 50, 60),
                       digitised = c(78, 62, 50, 40, 32, 26)) |>
  mutate(model = retentionAt(ivRun, day),
         ratio = model / digitised)
ggplot(ivRun, aes(day, Retention_mn54)) +
  geom_line() +
  geom_point(data = fig7, aes(day, digitised), inherit.aes = FALSE) +
  scale_y_log10() +
  labs(x = "Days", y = "Whole-body 54Mn retention (%)",
       title = "Replicates Figure 7 of Campbell 2023",
       subtitle = "Line = this implementation; points = digitised published curve") +
  theme_bw()
fig7 |>
  rename("Day" = day, "Figure 7 curve (%)" = digitised,
         "This implementation (%)" = model, "Ratio" = ratio) |>
  knitr::kable(digits = 3)
```

```{r human-fig7-assert}
stopifnot(
  # The paper's own claim for these data is "within a factor of 2"; our
  # agreement with its published curve is much tighter than that.
  all(fig7$ratio > 0.8), all(fig7$ratio < 1.25),
  # Retention must be monotonically decreasing and mass-balanced.
  all(diff(ivRun$Retention_mn54) <= 1e-8),
  abs(tail(ivRun$Mn54_balance, 1) - 1) < 1e-6
)
```

## Oral tracer and the published sensitivity fan, replicating Figure 9

Figure 9's caption prints all three (`FDIETUP`, `KBILEC`) pairs used to draw the
solid line and the two dashed sensitivity curves. That makes this a
**zero-parameter gate**: the inputs are printed, so the three output curves are
fully determined and nothing can be tuned.

```{r human-fig9}
fanSpec <- tibble::tribble(
  ~label,        ~f_diet_uptake, ~cl_bile, ~d10, ~d30, ~d60,
  "solid",        0.06,           0.051,    5.0,  1.80, 1.30,
  "upper dashed", 0.18,           0.017,    10.0, 4.50, 4.00,
  "lower dashed", 0.02,           0.153,    0.7,  0.25, 0.15
)
fanRuns <- lapply(seq_len(nrow(fanSpec)), function(i) {
  r <- tracerRun("oral", fanSpec$f_diet_uptake[i], fanSpec$cl_bile[i])
  r$label <- fanSpec$label[i]
  r$day <- r$time / 24
  r
})
fanAll <- dplyr::bind_rows(fanRuns)
ggplot(fanAll, aes(day, Retention_mn54, linetype = label)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Days", y = "Whole-body 54Mn retention (%)", linetype = NULL,
       title = "Replicates Figure 9 of Campbell 2023",
       subtitle = "Oral 54Mn in a labelled meal, with the printed 3-fold opposed FDIETUP/KBILEC fan") +
  theme_bw()
fanCmp <- lapply(seq_along(fanRuns), function(i) {
  r <- fanRuns[[i]]
  data.frame(label = fanSpec$label[i], day = c(10, 30, 60),
             digitised = c(fanSpec$d10[i], fanSpec$d30[i], fanSpec$d60[i]),
             model = retentionAt(r, c(10, 30, 60)))
}) |>
  dplyr::bind_rows() |>
  mutate(fold_error = pmax(model / digitised, digitised / model))
fanCmp |>
  rename("Curve" = label, "Day" = day, "Figure 9 (%)" = digitised,
         "This implementation (%)" = model, "Fold error" = fold_error) |>
  knitr::kable(digits = 3)
```

```{r human-fig9-assert}
meanFold <- exp(mean(abs(log(fanCmp$model / fanCmp$digitised))))
c(mean_fold_error = signif(meanFold, 4))
stopifnot(
  # Zero free parameters: all three input pairs are printed in the caption.
  meanFold < 1.6,
  max(fanCmp$fold_error) < 2.2,
  # The fan must be correctly ORDERED at every time point: raising FDIETUP and
  # lowering KBILEC together must raise retention, and vice versa.
  all(with(fanCmp[fanCmp$label == "upper dashed", ], model) >
        with(fanCmp[fanCmp$label == "solid", ], model)),
  all(with(fanCmp[fanCmp$label == "solid", ], model) >
        with(fanCmp[fanCmp$label == "lower dashed", ], model))
)
```

The day-10 retention of each curve sits close to its own `FDIETUP` (3.9% for
0.06, 11.2% for 0.18, 0.9% for 0.02). That is the structural signature of this
model: by day 10 the unabsorbed fraction has cleared the gut entirely, so
whole-body retention is essentially the absorbed fraction decaying slowly. It is
also the observation that identifies the human `KGI` -- see the Errata.

# Mass-balance and flux summary

```{r mass-balance-summary}
tibble::tribble(
  ~Check, ~Value,
  "Monkey, background diet at steady state", signif(mkLast$Mn_balance, 10),
  "Monkey, end of 90-day 1.5 mg/m3 inhalation", signif(mkPeak$Mn_balance, 10),
  "Human, background diet at steady state", signif(huLast$Mn_balance, 10),
  "Human, 54Mn after iv tracer (day 65)", signif(tail(ivRun$Mn54_balance, 1), 10),
  "Human, 54Mn after oral tracer (day 65)", signif(tail(fanRuns[[1]]$Mn54_balance, 1), 10)
) |>
  rename("Balance (should be 1)" = Value) |>
  knitr::kable()
```

```{r mass-balance-assert}
stopifnot(
  abs(mkLast$Mn_balance - 1) < 1e-6,
  abs(mkPeak$Mn_balance - 1) < 1e-6,
  abs(huLast$Mn_balance - 1) < 1e-6,
  abs(tail(ivRun$Mn54_balance, 1) - 1) < 1e-6,
  abs(tail(fanRuns[[1]]$Mn54_balance, 1) - 1) < 1e-6
)
```

Every manganese atom entering by diet or inhalation is accounted for either in
the body or in cumulative faecal excretion, in both species, at steady state and
under a cycling inhalation exposure, and for both the bulk and tracer systems.
Because this is an identity of the ODE system rather than a fitted quantity, it
goes red on almost any structural transcription error -- a dropped flux, a sign
error, or a term written into the wrong compartment.

# Assumptions and deviations

## Errata: three corrections to the published parameter tables

Each was decided by scoring both readings against the paper's own published
figures, not by preference.

**1. Table 2's `FDEPNO` / `FDEPNR` airflow-split factors are transposed.**
Table 2 prints `FDEPNO = 0.4313*0.91` and `FDEPNR = 0.4313*0.09` for the monkey
(and `0.3762*0.95` / `0.3762*0.05` for the human), i.e. the larger share of nasal
deposition to the olfactory region. The Methods say the opposite: "the MPPD
(ver. 3.04) head deposition was allocated to the respiratory and olfactory
region of the nose according to an airflow allocation of **9% olfactory, 91%
respiratory** (Kepler et al., 1998)". Using Table 2 as printed puts the
olfactory bulb at 21.1 ug/g against the published Figure 3J curve of 3.0
(**+602%**); using the Methods split gives 2.87 (**-4.4%**), and the median
error across all nine Figure 3 tissues improves from 9.3% to 4.3%. Both files
use the Methods split. Physiologically the olfactory epithelium receives only a
small fraction of nasal airflow, which is consistent with the Methods.

**2. Table 2's human `KGI` of 0.0026 /h is not the value the model uses; 0.06 /h
is.** At 0.0026 /h the gut-lumen half-life is 267 h, so unabsorbed oral tracer
is still in the gut at day 10 and Figure 9 retention comes out at ~54% instead
of the published ~5%; the printed three-curve sensitivity fan also collapses,
because retention is then dominated by gut contents rather than by `FDIETUP`. At
0.06 /h (the monkey value, half-life 11.6 h) all nine Figure 9 anchors land at a
1.15x mean fold error against 9.62x. The independent structural argument is in
the section above: Figure 9's day-10 retention tracks each curve's `FDIETUP`
directly, which is only possible once the gut has emptied. The human file uses
0.06 /h.

**3. The supplement's tracer mass-balance term double-integrates.** The listing
has `dt(XLOSS) = XFECES`, but `XFECES` is already the cumulative faecal tracer
state (`dt(XFECES) = XRFECES`); the bulk-manganese mirror is `dt(LOSS) = RFECES`,
a rate. As printed, the authors' own `balance54` diagnostic -- documented as
"value close to 1 means the model is mass balanced for 54Mn" -- cannot work.
Both files use the cumulative faecal tracer directly, which makes tracer mass
balance exactly 1 as intended.

## Other discrepancies, recorded but not acted on

* **Monkey `FDIETUP`.** Table 2 gives 2.00E-03; the Results text for the Dastur
  (1971) simulation says "the baseline dietary uptake (0.0002) was used". The
  files use Table 2's 0.002, following the rule that a printed parameter table
  has authority over prose.
* **Monkey body weight.** Table 1 lists 5.0 kg as the generic monkey reference,
  while the Results state that "the body weight of 2.5 kg corresponded to the
  average weight of the monkeys in Dorman et al. (2006a)" (and Dastur's animals
  also averaged 2.5 kg). Table 1's footnote a directs the use of study-specific
  weights where reported, so the monkey file defaults to 2.5 kg.
* **Human female liver volume.** Table 1 prints "0.233" as the female liver
  fraction of body weight, a decimal-point typo for 0.0233 (the male value is
  0.0247 and every other female entry is within 15% of the male one). The file
  uses the male reference values throughout, so the typo is inert here.
* **Dissociation half-lives.** The Discussion describes the dissociation rate
  constants as "equivalent to half-lives of 4-5 min"; from the printed `ka` and
  `KD` they are 9.1-11.4 min (tabulated above). No model parameter depends on
  the sentence.
* **`KD` in uM.** The Discussion prints "0.37 or 0.46 uM (20 or 25 ug/l)". Using
  the manganese molecular weight of 54.938 g/mol, 25 ug/L gives 0.455 uM, which
  rounds to the printed 0.46; 20 ug/L gives 0.364 uM, which rounds to 0.36, so
  the printed 0.37 is a one-in-the-last-digit round up. The files store the ug/L
  values from Table 2, so nothing downstream is affected.
* **Absolute human plasma manganese (Figure 10).** The model reproduces which
  dietary period gives the lowest and highest plasma manganese, and the shape of
  the response, but runs about 1.37-fold below the observed period means across
  all six periods (worst case 1.78-fold, in the 3.79 mg/day period, where the
  model's body pool equilibrates more slowly than the data over an 11-day
  window). Note that Figure 10 is the only human comparison that pins the
  ABSOLUTE steady-state level: the Figure 7 and 9 tracer comparisons are
  retention ratios, which constrain the clearance and distribution rates but
  cancel the absolute scale. The offset is therefore not inconsistent with those
  fits. No parameter was adjusted to reduce it.
* **Figure 10 x-axis.** The caption gives dietary periods spanning days 1-125,
  but the published panel's x-axis ends near day 110 and reports one plasma mean
  per period rather than per sampling day. Each observed mean is therefore placed
  at its period midpoint here. A point-by-point comparison against specific
  sampling days is not possible from the published caption.
* **`KM` units.** Table 2 labels the biliary-induction `KM` as ug/l, but the
  supplement compares it against `CART`, which is arterial ug/kg divided by
  1000, i.e. ug/g. The ug/g reading is the one consistent with the code and is
  what both files implement.
* **Brain blood flow.** The supplement computes brain flow as
  `QBRNC * QC * ((VSTMC + VOFBC + VCBMC + VPITC) / VBRNC)`. Dividing by `VBRNC`
  is dimensionally odd, because the region fractions are already fractions of
  brain. The printed form is nevertheless retained, because it is what
  reproduces Figure 3: it puts the globus pallidus, pituitary and cerebellum
  within 8-13% of the published curves, whereas dropping the divisor moves them
  to 42-61% off.
* **Vestigial state.** The listing's `ABILE` state has
  `RABILE = RBILE - QBILE*CBILE` with `CBILE = RBILE/QBILE`, so its derivative
  is identically zero. It is omitted; the bile *concentration* output it was
  presumably meant to support is retained as `Cbile` in the monkey file.

## Parameters not reported anywhere on disk

Five constants sit in the supplement's `# Dosing Controls` block, which the
supplement introduces with "Note: parameters are set in simulation scripts" --
they are scenario inputs, not fitted parameters -- and no value for any of them
appears in the paper, its supplement, or any other on-disk source. **No value
has been invented for any of them.**

```{r unreported, echo = FALSE}
tibble::tribble(
  ~Parameter, ~`Role`, ~`Consequence`,
  "INFAC (f_intake_diet)", "converts a ppm diet to an intake rate", "the ppm dietary route is inert by default; use diet_mn_mgd instead (the route the paper specifies numerically)",
  "IPFRAC, KIPSLOW (kip)", "intraperitoneal tracer sub-model", "Figure 4A (Dastur 1971 ip) is not reproduced",
  "KSBQ (ksc)", "subcutaneous absorption rate", "Figure 5A (Newland 1987 sc infusion) is not reproduced",
  "SANEWLAND", "scales predicted faecal tracer concentration", "Figure 5 faecal concentrations are not reproduced"
) |>
  knitr::kable()
```

Consequences for validation scope:

* The **model itself is complete** -- every physiological (Table 1) and
  chemical-specific (Table 2) parameter is reported, which is why the Figure 3,
  6, 7, 9 and 10 reproductions above work.
* The **human model is fully validatable**: its diet is specified in mg/day and
  Figures 7, 9 and 10 use only fully specified routes.
* The **monkey inhalation** validation -- the paper's primary monkey dataset and
  the actual parameter-optimisation target -- is reproduced, with the dietary
  rate calibrated to the reported Figure 3F baseline blood manganese and
  labelled as a calibrated scenario input.
* The monkey **intraperitoneal, subcutaneous and ppm-diet** tracer routes
  (Figures 4A and 5) are implemented structurally but are **not validated here**.
  Schroeter et al. (2011), the upstream source of `INFAC`, is not on disk.

## Other assumptions

* **Burn-in.** Both files start all states at zero, as the supplement's listing
  does. Every simulation above therefore begins with a burn-in to steady state
  on the background diet (1500 days for the monkey, 4000 for the human, both
  verified to have converged). A user who skips the burn-in will get a
  manganese-depleted animal.
* **No variability.** The publication reports no inter-individual variability
  and no residual error, so neither is encoded. Every simulation here is a
  typical-value solve and none draws random numbers, so nothing in this vignette
  depends on a seed or on the solver's thread count.
* **Figure digitisation.** The Figure 3, 6, 7, 9 and 10 comparison values were
  read off the published raster figures, so they carry digitisation error of
  roughly a few percent (more on the log-scale retention panels). The assertion
  thresholds above are set accordingly, and no model parameter was derived from
  a figure.
* **Species labels.** The supplement's code names the globus-pallidus states
  `ST`/`STM` ("striatum"), a label inherited from the Yoon (2019) rat model. The
  compartment is the globus pallidus: its volume fraction is Table 1's "Globus
  Pallidus" row, supplementary Table S1 lists caudate, putamen and globus
  pallidus separately, and the paper's sensitivity endpoint is explicitly the
  globus pallidus. The canonical name `brain_globus_pallidus` is used.
