---
title: "Brilaroxazine / RP5063 (Cantillon 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Brilaroxazine / RP5063 (Cantillon 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  error = FALSE
)
```

```{r libraries, message = FALSE, warning = FALSE}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(ggplot2)
```

This vignette validates `Cantillon_2018_brilaroxazine`, the joint population
PK / PANSS PD model for RP5063 (brilaroxazine) reported in

> Cantillon M, Ings R, Prakash A, Bhat L (2018). A population pharmacokinetic
> and pharmacodynamic analysis of RP5063 phase 2 study data in patients with
> schizophrenia or schizoaffective disorder. *European Journal of Drug
> Metabolism and Pharmacokinetics* 43(5):573-585.
> [doi:10.1007/s13318-018-0472-z](https://doi.org/10.1007/s13318-018-0472-z)

```{r load-model}
mod <- modellib("Cantillon_2018_brilaroxazine")
# Typical-value (inter-individual variability zeroed) version, used for the
# closed-form and structural checks below.
mod_typ <- rxode2::zeroRe(mod)
mod
```

# Population

The analysis dataset is the RP5063 arm of the phase 2 REFRESH trial
(NCT01490086), an in-patient, international, randomised, double-blind,
placebo-controlled study. Of 234 randomised patients, **175** received RP5063
at 15, 30 or 50 mg orally once daily for 28 days, dosed after an overnight
fast and 1 h before breakfast; only those 175 entered the population analysis.

Baseline characteristics are from Cantillon 2018 Table 1.

```{r population}
pop <- readModelDb("Cantillon_2018_brilaroxazine") |> attr("population")
tibble::tibble(Field = names(pop), Value = vapply(pop, function(x) {
  paste(if (is.null(names(x))) x else paste0(names(x), " ", x, "%"), collapse = "; ")
}, character(1))) |>
  knitr::kable(caption = "Population metadata (Cantillon 2018 Table 1 and Sect. 2.1, 3.1).")
```

Five plasma samples per patient were drawn (pre-dose at baseline and on days 1,
8, 22 and 28, allocated across four time-blocks) out to 220 h after the last
dose, deliberately spanning both non-steady-state and steady-state conditions
so that `Vc/F` and `Cl/F` could be separated. Total PANSS was collected
pre-dose on day 1 and at least 2 h post-dose on days 4, 8, 15, 22 and 28.

Note a small internal inconsistency in the source: Sect. 3.1 states the cohort
was 80% male / 20% female, but the Table 1 counts give 133 male and 42 female,
i.e. 24% female. The Table 1 counts are used in the model metadata.

# Source trace

Every structural equation and every `ini()` value, with its location in the
source.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Quantity, ~Value, ~`Source location`,
  "One-compartment, first-order absorption, lag time", "structure", "Eq. 1; Sect. 3.2",
  "Cl/F", "5.11 L/h (SE 0.11)", "Table 2",
  "Vc/F (at BMI 23.01)", "328 L (SE 31.4)", "Table 2",
  "ka", "0.42 1/h (SE 0.17)", "Table 2",
  "t lag", "0.41 h (SE 0.02)", "Table 2",
  "BMI power on Vc (c1)", "0.90 (SE 0.36)", "Table 2 row 'c'; Eq. 18",
  "BMI centring constant", "23.01 kg/m2", "Eq. 18 (cohort mean; Eq. 3 divides by mean(cov))",
  "Var(eta Cl), Cov, Var(eta Vc)", "0.16, 0.04, 0.28", "Table 2",
  "Var(eta ka)", "2.09 (SE 6.63)", "Table 2",
  "Proportional residual (sigma_1^2)", "0.07 -> SD 0.2646", "Table 2 row 'r 1 2'; Sect. 3.2 calls it a CV",
  "PD driver = cumulative AUC", "predictor selection", "Eq. 11; Sect. 3.4",
  "Emax form E0 + Emax x/(x + x50)", "structure", "Eq. 12",
  "E0", "87.3 pu (SE 0.711)", "Table 3",
  "Emax", "-31.6 pu (SE 4.05)", "Table 3",
  "AUC50", "89.6 ug*h/mL (SE 30.1)", "Table 3",
  "Moldova-site effect on Emax (c1)", "-1.93 (SE 0.535)", "Table 3 row 'c'; Sect. 3.4",
  "Var(eta E0) additive", "164", "Table 3; Eq. 14",
  "Var(eta Emax) additive", "464", "Table 3; Eq. 15",
  "Var(eta AUC50) log-normal", "0.476", "Table 3; Eq. 16",
  "Additive PANSS residual (sigma^2)", "47.1 -> SD 6.863", "Table 3 row 'r 2'",
  "Placebo term omitted", "rejected", "Sect. 3.4, 4 (objective function increased)"
) |>
  knitr::kable(caption = "Source trace for Cantillon_2018_brilaroxazine.")
```

# Closed-form checks

These are exact consequences of the Table 2 / Table 3 point estimates and do
not depend on any simulated cohort, so they are asserted tightly.

```{r closed-form}
cl <- 5.11
vc <- 328
thalf <- log(2) * vc / cl
emax_moldova <- -31.6 * (1 + (-1.93) * 1)

tibble::tribble(
  ~Check, ~Paper, ~Model,
  "Terminal half-life (h)", 44.5, round(thalf, 2),
  "Typical Emax at Moldova sites (pu)", 29.4, round(emax_moldova, 2),
  "Effect at cumulative AUC = AUC50 (pu)", -15.8, round(-31.6 * 89.6 / (89.6 + 89.6), 2)
) |>
  knitr::kable(caption = "Closed-form quantities the paper states explicitly.")

stopifnot(
  # Sect. 3.3 / Key Points: "a calculated half-life of 44.5 h", derived from
  # Vc/F and Cl/F. Pure arithmetic on Table 2, so a tight bound is correct.
  abs(thalf - 44.5) < 0.05,
  # Sect. 3.4: c1 = -1.93 "corresponds to an average positive Emax of 29.4 (pu)".
  abs(emax_moldova - 29.4) < 0.05
)
```

# Virtual cohort

Three arms of 100 subjects each (well under the 200-per-arm cap), matching the
15 / 30 / 50 mg arms of REFRESH. BMI is drawn to reproduce the Table 1 arm
means and SDs (23.2 / 22.3 / 23.4 kg/m^2, SD 4.2 / 4.6 / 3.3) and truncated to
a physiologically plausible adult range. The Moldova-site indicator is set to 0
for the main cohort so the base dose-response relationship is reproduced; the
site effect is examined separately below.

```{r cohort}
rxode2::rxSetSeed(20260919)
set.seed(20260919)

n_arm <- 100L
arms <- tibble::tibble(
  treatment = c("15 mg", "30 mg", "50 mg"),
  dose_mg = c(15, 30, 50),
  bmi_mean = c(23.2, 22.3, 23.4),
  bmi_sd = c(4.2, 4.6, 3.3)
)

cohort <- arms |>
  rowwise() |>
  do({
    a <- .
    tibble::tibble(
      treatment = a$treatment,
      dose_mg = a$dose_mg,
      BMI = pmin(pmax(rnorm(n_arm, a$bmi_mean, a$bmi_sd), 15), 45),
      REGION_MOLDOVA = 0
    )
  }) |>
  ungroup() |>
  mutate(id = row_number())

summary_bmi <- cohort |>
  group_by(treatment) |>
  summarise(n = n(), BMI_mean = round(mean(BMI), 2), BMI_sd = round(sd(BMI), 2),
            .groups = "drop")
knitr::kable(summary_bmi, caption = "Simulated cohort BMI vs. Cantillon 2018 Table 1.")
```

```{r event-builder}
# Build event tables as plain data frames. Both endpoints (Cc and PANSS) are
# declared in the model, so observation rows must name an observable; PANSS and
# auc_central are still returned as columns at those rows.
make_events <- function(cohort, times, n_doses, ii = 24) {
  doses <- cohort |>
    tidyr::expand_grid(dose_idx = seq_len(n_doses)) |>
    transmute(id, treatment, BMI, REGION_MOLDOVA,
              time = (dose_idx - 1) * ii, amt = dose_mg,
              evid = 1L, cmt = "depot")
  obs <- cohort |>
    tidyr::expand_grid(time = times) |>
    transmute(id, treatment, BMI, REGION_MOLDOVA,
              time, amt = NA_real_, evid = 0L, cmt = "Cc")
  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}
```

# Pharmacokinetics: 28-day once-daily regimen

```{r sim-md, message = FALSE, warning = FALSE}
rxode2::rxSetSeed(20260919)

times_md <- sort(unique(c(seq(0, 672, by = 6), seq(0, 24, by = 2))))
ev_md <- make_events(cohort, times_md, n_doses = 28)

sim_md <- rxode2::rxSolve(mod, ev_md, keep = c("treatment", "BMI")) |>
  as.data.frame()
```

Cantillon 2018 Figure 3 shows the population bootstrap prediction with 95%
point-wise bands for each dose. The corresponding simulated prediction
intervals are below.

```{r fig-pk, fig.cap = "Simulated RP5063 plasma concentration over the 28-day regimen (median and 90% prediction interval). Replicates the shape of Figure 3 of Cantillon 2018."}
pk_bands <- sim_md |>
  group_by(treatment, time) |>
  summarise(med = median(Cc), lo = quantile(Cc, 0.05), hi = quantile(Cc, 0.95),
            .groups = "drop")

ggplot(pk_bands, aes(time / 24, med)) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = treatment), alpha = 0.2) +
  geom_line(aes(colour = treatment), linewidth = 0.7) +
  labs(x = "Time (days)", y = "RP5063 plasma concentration (ng/mL)",
       colour = "Dose", fill = "Dose") +
  theme_bw()
```

Steady state is approached over roughly the first 5 days, consistent with the
phase 1 multiple-dose observation quoted in Sect. 1 ("steady state was
approached after 120 h of daily dosing").

## Dose linearity

Sect. 3.3 reports pairwise *t*-tests that failed to reject equality of `Cl/F`
across the three dose arms, i.e. linear PK over 15-50 mg. The model has no
non-linear term, so exposure must be exactly dose-proportional.

Proportionality is a property of each individual, so it is asserted on the
typical-value solve, where the three arms share one parameter vector. (Medians
across three *independently drawn* arms are not exactly proportional, because
each arm draws its own BMI and random effects.)

```{r linearity}
lin_typ <- tibble::tibble(treatment = c("15 mg", "30 mg", "50 mg"),
                          dose_mg = c(15, 30, 50), BMI = 23.01,
                          REGION_MOLDOVA = 0) |>
  mutate(id = row_number())

sim_lin <- rxode2::rxSolve(mod_typ, make_events(lin_typ, c(0, 672), n_doses = 28),
                           keep = "treatment") |>
  as.data.frame() |>
  filter(time == 672) |>
  left_join(lin_typ |> select(id, dose_mg), by = "id") |>
  transmute(treatment, `Dose (mg)` = dose_mg,
            `Cumulative AUC (ug*h/mL)` = auc_central,
            `AUC per mg` = auc_central / dose_mg)

knitr::kable(sim_lin, digits = 4,
             caption = "Typical-value cumulative AUC at day 28 scales exactly with dose.")

stopifnot(
  # Deterministic consequence of a linear model: solver noise only.
  diff(range(sim_lin$`AUC per mg`)) / mean(sim_lin$`AUC per mg`) < 1e-6
)
```

In the stochastic cohort the per-arm medians are close but not identical,
reflecting the independent covariate and random-effect draws in each arm:

```{r linearity-cohort}
sim_md |>
  filter(time == max(time)) |>
  group_by(treatment) |>
  summarise(`Median cumulative AUC (ug*h/mL)` = median(auc_central),
            .groups = "drop") |>
  mutate(`Dose (mg)` = c(15, 30, 50),
         `AUC per mg` = `Median cumulative AUC (ug*h/mL)` / `Dose (mg)`) |>
  knitr::kable(digits = 3,
               caption = "Stochastic cohort medians (descriptive).")
```

## Body mass index affects volume but not average steady-state exposure

Sect. 4 makes an explicit, directly testable claim:

> Since BMI only affected volume of distribution, this covariate should not
> affect the average steady-state plasma levels of RP5063, since clearance did
> not change.

```{r bmi-check}
bmi_grid <- tibble::tibble(BMI = c(18, 23.01, 30, 35)) |>
  mutate(id = row_number(), treatment = "30 mg", dose_mg = 30, REGION_MOLDOVA = 0)

ev_bmi <- make_events(bmi_grid, seq(0, 672, by = 1), n_doses = 28)
sim_bmi <- rxode2::rxSolve(mod_typ, ev_bmi, keep = c("BMI")) |> as.data.frame()

bmi_tab <- sim_bmi |>
  filter(time >= 648) |>
  group_by(BMI) |>
  summarise(`Cavg day 28 (ng/mL)` = mean(Cc),
            `Cmax day 28 (ng/mL)` = max(Cc), .groups = "drop") |>
  mutate(`Vc/F (L)` = 328 * (BMI / 23.01)^0.90,
         `t1/2 (h)` = log(2) * `Vc/F (L)` / 5.11) |>
  select(BMI, `Vc/F (L)`, `t1/2 (h)`, `Cavg day 28 (ng/mL)`, `Cmax day 28 (ng/mL)`)

knitr::kable(bmi_tab, digits = 2,
             caption = "BMI moves Vc/F and half-life substantially, but leaves average steady-state concentration unchanged (Cantillon 2018 Sect. 4).")

stopifnot(
  # Volume nearly doubles across the BMI range examined...
  max(bmi_tab$`Vc/F (L)`) / min(bmi_tab$`Vc/F (L)`) > 1.7,
  # ...while average steady-state concentration moves by well under 1%.
  diff(range(bmi_tab$`Cavg day 28 (ng/mL)`)) /
    mean(bmi_tab$`Cavg day 28 (ng/mL)`) < 0.01
)
```

# PKNCA validation

The paper reports one NCA-comparable quantity, the calculated terminal
half-life of 44.5 h. The typical-value (inter-individual variability zeroed)
single-dose profile is used for the gate, because the model's IIV on `ka`
(variance 2.09, i.e. a 266% CV) puts a meaningful fraction of simulated
subjects into flip-flop kinetics, where a terminal-slope half-life reflects
absorption rather than elimination. The stochastic cohort is reported
descriptively alongside.

```{r nca-typical, message = FALSE, warning = FALSE}
typ <- tibble::tibble(
  treatment = c("15 mg", "30 mg", "50 mg"), dose_mg = c(15, 30, 50),
  BMI = 23.01, REGION_MOLDOVA = 0
) |> mutate(id = row_number())

times_sd <- c(0, 0.5, 1, 2, 3, 4, 6, 8, 10, 12, 16, 20, 24, 30, 36, 48, 60, 72,
              96, 120, 168, 216, 264, 336, 408, 504)
ev_sd <- make_events(typ, times_sd, n_doses = 1)

sim_sd <- rxode2::rxSolve(mod_typ, ev_sd, keep = c("treatment")) |>
  as.data.frame()

conc_sd <- sim_sd |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)
conc_sd <- bind_rows(
  conc_sd,
  conc_sd |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

dose_sd <- ev_sd |> filter(evid == 1) |> select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(conc_sd, Cc ~ time | treatment + id,
                             concu = "ng/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_sd, amt ~ time | treatment + id, doseu = "mg")

intervals <- data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                        aucinf.obs = TRUE, half.life = TRUE, lambda.z = TRUE)

nca_typ <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
```

The reference column below carries the paper's calculated half-life of 44.5 h
(Sect. 3.3 and Key Points) and the AUC0-inf implied by the paper's own `Cl/F`
of 5.11 L/h (Table 2), i.e. dose / (Cl/F), converted to ng*h/mL. The latter is
derived from Table 2 rather than tabulated as an NCA result in the paper.

```{r nca-compare}
published <- tibble::tibble(
  treatment = c("15 mg", "30 mg", "50 mg"),
  aucinf.obs = c(15, 30, 50) / 5.11 * 1000,
  half.life = 44.5
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_typ,
  reference = published,
  by = "treatment",
  units = c(aucinf.obs = "ng*h/mL", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(cmp, caption = paste(
  "Simulated (typical-value) vs. published NCA.",
  "* marks rows differing from the reference by more than 20%."
))
```

```{r nca-assert}
nca_wide <- as.data.frame(nca_typ$result) |>
  filter(PPTESTCD %in% c("half.life", "aucinf.obs")) |>
  tidyr::pivot_wider(id_cols = treatment, names_from = PPTESTCD,
                     values_from = PPORRES)

stopifnot(
  # Deterministic typical-value profile against the paper's closed-form
  # half-life, so a tight bound is appropriate here.
  all(abs(nca_wide$half.life - 44.5) < 0.5),
  # AUC0-inf must recover dose / (Cl/F) to within trapezoidal error.
  all(abs(nca_wide$aucinf.obs / (c(15, 30, 50) / 5.11 * 1000) - 1) < 0.02)
)
```

```{r nca-cohort, message = FALSE, warning = FALSE}
rxode2::rxSetSeed(20260919)

ev_sd_pop <- make_events(cohort, times_sd, n_doses = 1)
sim_sd_pop <- rxode2::rxSolve(mod, ev_sd_pop, keep = c("treatment")) |>
  as.data.frame()

conc_pop <- sim_sd_pop |> filter(!is.na(Cc)) |> select(id, time, Cc, treatment)
conc_pop <- bind_rows(
  conc_pop,
  conc_pop |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

nca_pop <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_pop, Cc ~ time | treatment + id,
                   concu = "ng/mL", timeu = "h"),
  PKNCA::PKNCAdose(ev_sd_pop |> filter(evid == 1) |>
                     select(id, time, amt, treatment),
                   amt ~ time | treatment + id, doseu = "mg"),
  intervals = intervals
))

as.data.frame(nca_pop$result) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "half.life", "aucinf.obs")) |>
  group_by(treatment, PPTESTCD) |>
  summarise(Median = median(PPORRES, na.rm = TRUE),
            `P10` = quantile(PPORRES, 0.10, na.rm = TRUE),
            `P90` = quantile(PPORRES, 0.90, na.rm = TRUE), .groups = "drop") |>
  mutate(Parameter = nlmixr2lib::ncaParamLabel(PPTESTCD)) |>
  select(Parameter, treatment, Median, P10, P90) |>
  knitr::kable(digits = 2, caption = paste(
    "Stochastic cohort NCA (descriptive). The wide half-life spread is driven",
    "by the 266% CV on ka, which pushes slowly-absorbing subjects into",
    "flip-flop kinetics."
  ))
```

# Pharmacodynamics: PANSS versus cumulative exposure

Model selection (Sect. 3.4) found cumulative AUC to be the best of the four
candidate predictors (instantaneous concentration, effect-compartment
concentration, daily average concentration, cumulative AUC), and a placebo
term made the fit worse and was dropped.

```{r fig-panss-auc, fig.cap = "Predicted total PANSS as a function of cumulative RP5063 AUC. Replicates the model fit line of Figure 4 of Cantillon 2018."}
auc_grid <- tibble::tibble(auc = seq(0, 400, length.out = 200)) |>
  mutate(PANSS = 87.3 + (-31.6) * auc / (auc + 89.6))

ggplot(auc_grid, aes(auc, PANSS)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = 89.6, linetype = "dotted") +
  annotate("text", x = 89.6, y = 86, label = "AUC50 = 89.6", hjust = -0.05,
           size = 3) +
  labs(x = "Cumulative AUC (ug*h/mL)", y = "Total PANSS") +
  theme_bw()
```

## Dose-response relationship (Figure 6)

Figure 6 of the paper plots predicted total PANSS against daily dose, with a
solid line for the base model and a dashed line for the model carrying the
geographic-site effect on Emax. Both are reproduced by solving the typical-value
model for 28 days over a dose grid.

```{r fig-dose-response, message = FALSE, warning = FALSE, fig.cap = "Predicted total PANSS at day 28 versus daily dose, with and without the Moldova study-site effect on Emax. Replicates Figure 6 of Cantillon 2018."}
dose_grid <- tidyr::expand_grid(
  dose_mg = seq(2.5, 60, by = 2.5),
  REGION_MOLDOVA = c(0, 1)
) |>
  mutate(id = row_number(), BMI = 23.01,
         treatment = ifelse(REGION_MOLDOVA == 1, "With Moldova site effect",
                            "Base model"))

ev_dr <- make_events(dose_grid, c(0, 672), n_doses = 28)
sim_dr <- rxode2::rxSolve(mod_typ, ev_dr,
                          keep = c("treatment", "REGION_MOLDOVA")) |>
  as.data.frame() |>
  filter(time == 672) |>
  left_join(dose_grid |> select(id, dose_mg), by = "id")

ggplot(sim_dr, aes(dose_mg, PANSS, linetype = treatment)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 87.3, colour = "grey60", linetype = "dotted") +
  annotate("rect", xmin = 5, xmax = 30, ymin = -Inf, ymax = Inf, alpha = 0.08,
           fill = "steelblue") +
  labs(x = "Daily dose (mg)", y = "Predicted total PANSS at day 28",
       linetype = NULL,
       caption = "Shaded band: the 5-30 mg range the paper identifies as clinically relevant.") +
  theme_bw() +
  theme(legend.position = "bottom")
```

The paper concludes (Sect. 3.4, Sect. 5) that "daily doses between 5 and 30 mg
should adequately describe a clinically relevant dose/efficacy relationship".
The simulated curve reproduces that: most of the attainable PANSS improvement
is realised by 30 mg, and the curve flattens above it.

```{r dose-response-check}
base_dr <- sim_dr |> filter(REGION_MOLDOVA == 0) |> arrange(dose_mg)
get_change <- function(d) {
  base_dr$PANSS[which.min(abs(base_dr$dose_mg - d))] - 87.3
}

dr_tab <- tibble::tibble(
  `Daily dose (mg)` = c(5, 15, 30, 50),
  `PANSS change at day 28 (pu)` = round(vapply(c(5, 15, 30, 50), get_change, 0), 2)
)
knitr::kable(dr_tab, caption = "Predicted PANSS change from baseline at day 28.")

stopifnot(
  # Monotonically increasing improvement with dose.
  all(diff(base_dr$PANSS) < 0),
  # The 5-30 mg window captures most of the change realised by 50 mg. The
  # changes are negative (PANSS improves by falling), so compare magnitudes.
  abs(get_change(30) - get_change(5)) >
    2 * abs(get_change(50) - get_change(30)),
  # ...and the Moldova arm moves PANSS in the opposite direction.
  all(sim_dr$PANSS[sim_dr$REGION_MOLDOVA == 1 & sim_dr$dose_mg > 5] > 87.3)
)
```

## Comparison with the observed trial arm means

The REFRESH primary-endpoint analysis (quoted in Sect. 1) reported mean (SE)
PANSS changes of -20.23 (2.65), -15.42 (2.04) and -19.21 (2.39) for the 15, 30
and 50 mg arms.

```{r arm-means}
sim_panss <- sim_md |>
  filter(time == 672) |>
  group_by(treatment) |>
  summarise(`Model median change (pu)` = round(median(PANSS) - 87.3, 2),
            .groups = "drop") |>
  mutate(`Observed change (pu)` = c(-20.23, -15.42, -19.21),
         `Observed SE` = c(2.65, 2.04, 2.39))

knitr::kable(sim_panss, caption = paste(
  "Model-predicted vs. observed PANSS change at day 28. The Emax model is",
  "monotone in dose by construction and does not reproduce the non-monotone",
  "ordering of the observed arm means."
))
```

The model is monotone in dose, whereas the observed arm means are not
(the 30 mg arm improved least). This is a property of the published model, not
of the encoding: Cantillon 2018 fitted a single Emax curve through all
subject-level PANSS-versus-exposure pairs rather than to the arm means, and
Figures 4 and 5 show the corresponding scatter. No parameter has been adjusted
to improve this comparison.

# Assumptions and deviations

* **Variance versus standard deviation in Tables 2 and 3.** The tables label
  their random-effect rows "Var(g1)", "Cov(g1, g2)" and "sigma_1^2" / "sigma^2"
  (i.e. sigma^2), so the tabulated numbers are read as **variances**. Three
  independent checks agree:
  (a) the covariance row implies a correlation of
  `0.04 / sqrt(0.16 * 0.28) = 0.19` under the variance reading, against 0.89
  under a standard-deviation reading;
  (b) `Var(eta E0) = 164` gives an E0 standard deviation of 12.8 PANSS units,
  matching the observed per-arm baseline PANSS SDs of 13.3 / 13.4 / 14.9 in
  Table 1, whereas a standard-deviation reading would place 164 PANSS units of
  spread on a scale bounded at 30-210;
  (c) the empirical-Bayes `Cl/F` of Sect. 3.3, 5.17 +/- 0.24 L/h over 58
  subjects, is a mean +/- standard error, implying an EBE standard deviation of
  1.83 L/h and a CV of 35% - consistent with the 42% CV of `Var = 0.16` after
  shrinkage, but impossible under the 16% CV of a standard-deviation reading,
  since empirical-Bayes spread cannot exceed the true IIV.
* **Errata: the base-model variability percentages in Sect. 3.2 are
  inconsistent with Table 2.** Sect. 3.2 reports base-model inter-subject CVs
  of "14.90 and 21.40% for Cl/F and Vc/F" and an intra-subject CV of
  "approximately 8.0%". Under the variance reading adopted here the final-model
  values are 42%, 57% and 26%. The Sect. 3.2 figures are also internally
  suspicious: 14.90 is the same number as the Vc/F standard error quoted in the
  immediately preceding sentence. Because a covariate addition cannot inflate
  an IIV variance sevenfold, and because check (c) above rules the small values
  out directly, Table 2 is used. A reader who preferred the Sect. 3.2 reading
  would set `etalcl ~ 0.0222`, `etalvc ~ 0.0459` and `propSd <- 0.08`.
* **Supplementary appendices are not available.** Appendices A-D (sampling-time
  blocks, bioanalytical method, PD covariate stepwise table, concomitant
  medication list) could not be obtained when this model was built. None
  contains a parameter used by the model; the covariate-screen results they
  support are documented in `covariatesDataExcluded`.
* **`ka` is poorly identified.** Table 2 gives `Var(g3) = 2.09` with a standard
  error of 6.63, i.e. a relative standard error above 300%, and the point
  estimate of `ka` itself has an RSE of 40%. Sect. 4 attributes this to the
  absence of early post-dose sampling. Simulated single-subject absorption
  profiles are correspondingly variable and a small fraction show flip-flop
  kinetics; this is faithful to the published model, not an encoding artefact.
* **Cumulative, not steady-state, AUC.** The `auc_central` state integrates
  concentration from the first dose and therefore keeps rising for the whole
  28-day course. This is the paper's Eq. 11 predictor. It is not a
  dosing-interval AUC, and `AUC50 = 89.6 ug*h/mL` is not comparable to a
  steady-state AUC0-24 for this drug (which is about 3-10 ug*h/mL).
* **The Moldova site effect is a measurement artefact, per the authors.**
  Sect. 3.4 and Sect. 4 report that the site influenced only the PANSS
  endpoint, not the PK, and describe the data as an outlier. It is encoded
  because it is in the final model, but `REGION_MOLDOVA = 0` reproduces the
  paper's base dose-response relationship.
* **No placebo term.** Eq. 13's `alpha * t` placebo component was tested and
  rejected (the objective function increased); no `alpha` estimate is reported,
  so none is encoded.
* **No bioavailability parameter.** `Cl/F` and `Vc/F` are apparent oral values
  throughout; the paper reports no `F` and no intravenous reference arm, so
  absolute volume and clearance are not identifiable.
* **Drug naming.** The paper calls the compound RP5063 throughout; the model
  file uses the INN, brilaroxazine, per the library's generic-name convention.
* **Two-compartment structure not carried.** Sect. 3.2 reports that a
  two-compartment fit produced parameter correlations above 0.95 and no
  computable standard errors, so the authors selected the one-compartment
  model; Sect. 4 notes the distribution phase carries under 10% of total AUC.
  Only the published one-compartment model is encoded.

# Session info

```{r session-info}
sessionInfo()
```
