---
title: "Busulfan (Cao 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Busulfan (Cao 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r meta}
mod <- readModelDb("Cao_2025_busulfan")
cat(rxode2::rxode(mod)$reference)
```

* Article: <https://doi.org/10.3389/fphar.2025.1632588>
* Supplementary material (Text S1 body-size formulae, Tables S1-S3, **Text S2 =
  the final NONMEM control stream**):
  <https://www.frontiersin.org/articles/10.3389/fphar.2025.1632588/full#supplementary-material>

Cao 2025 is a semi-mechanistic population PK model for intravenous busulfan in
paediatric allogeneic haematopoietic cell transplantation (HCT). On top of a
conventional two-compartment disposition model it adds a third state: a
glutathione (GSH) pool, normalised to 1 at baseline, that multiplies the
busulfan elimination rate constant and is itself driven by busulfan metabolic
flux. Baseline glutathione S-transferase (GST) enzyme activity enters as a power
covariate on the coupling constant. The result is a clearance that changes over
a multi-day conditioning course.

This vignette does four things:

1. Reproduces Figure 3 (the GST-activity effect on the coupling constant)
   exactly.
2. Checks two closed-form structural identities (the maturation half-time and
   the dose/clearance identity when the GSH coupling is switched off).
3. Reproduces **all eleven rows of Table 3** -- the paper's own virtual-trial
   cumulative-exposure simulation -- which serves as an answer key because the
   paper publishes the complete covariate vector for every virtual subject.
4. Runs a PKNCA cross-check on the simulated profiles.

The Table 3 reproduction is also what settles a sign ambiguity in the published
model; see [Errata and interpretation](#errata-and-interpretation).

# Population

The model was developed from 636 whole-blood busulfan concentrations collected
prospectively from 65 paediatric HCT recipients at the Children's Hospital of
Fudan University (Shanghai, China) between August 2020 and November 2021 (Cao
2025 Methods 2.1, Table 1, Figure 1). All subjects received 0.8-1.2 mg/kg per
dose as a 2 h IV infusion every 6 h for 12 doses, with the dose band chosen by
body weight.

The cohort was split into a model-development set (55 subjects, 536 samples) and
a held-out evaluation set (10 subjects, 100 samples). In the development set the
median age was 1.4 years (range 0.2-14.1, with 21 subjects under 1 year), median
weight 9.9 kg (range 2.9-29.5), median height 76.0 cm (range 52.0-147.0), and
median fat-free mass 8.5 kg (range 2.5-28.0). 40 of 55 were male. Median
baseline GST enzyme activity was 9.2 nmol/min/mL (range 0.9-20.7) -- this median
is the centering constant used by the GST covariate model.

Sampling was at 2, 2.5, 3, 4 and 6 h after dose 1; pre-dose before doses 6 and
12; and at 2/4/8 h (Group A, n = 33) or 2/6/12 h (Group B, n = 32) after dose 12.
Those four sampling windows are the four inter-occasion-variability occasions.

Programmatic access to the same metadata:

```{r population-metadata}
str(mod()$population)
```

# Source trace

Every `ini()` entry in `inst/modeldb/specificDrugs/Cao_2025_busulfan.R` carries
an in-file comment naming its origin. Because the supplement publishes the final
NONMEM control stream (Text S2) *and* the paper publishes a parameter table
(Table 2), almost every value has two independent sources that agree. The table
below collects them.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` (CL) | 9.57 L/h | Table 2; Text S2 `$THETA(1)` |
| `lvc` (Vc) | 28.2 L | Table 2; Text S2 `$THETA(2)` |
| `lq` (Q) | 8.16 L/h | Table 2; Text S2 `$THETA(3)` |
| `lvp` (Vp) | 16.1 L | Table 2; Text S2 `$THETA(4)` |
| `ffat_cl` | 0.905 | Table 2; Text S2 `$THETA(5)` |
| `ffat_vc` | 0.687 | Table 2; Text S2 `$THETA(6)` |
| `tm50_mat` | 45.0 weeks | Table 2; Text S2 `$THETA(7)` |
| `hill_mat` | 1.11 | Table 2; Text S2 `$THETA(8)` |
| `sdep_gsh` | 0.00259 L/mg (fixed) | Table 2; Text S2 `$THETA(9)`; value carried from Langenhorst 2020 |
| `e_gst_sdep_gsh` | 0.28 | Table 2; Text S2 `$THETA(10)` |
| `etalcl` | 0.0539 | Table 2 BSV CL 23.2% (see Errata: the stream's `$OMEGA` prints 0.539) |
| `etalvc` | 0.0244 | Table 2 BSV Vc 15.6%; Text S2 `$OMEGA` record 2 |
| `etalvp` | 0.16 | Table 2 BSV Vp 40.0%; Text S2 `$OMEGA` record 3 |
| `etaiov_cl_1..4` | 0.0115 | Table 2 IOV on CL 10.7%; Text S2 `$OMEGA BLOCK(1)` + 3x `SAME` |
| `propSd` | 0.1114 | Text S2 `$SIGMA(1)` = 0.0124; Table 2 "Proportional 11.1%" |
| `addSd` | 0.01661 mg/L | Text S2 `$SIGMA(2)` = 276 (ng/mL)^2; Table 2 "Additional 16.6" |
| NFM = FFM + Ffat x (WT - FFM) | n/a | Supplementary Text S1 Eq. 4; Text S2 `$PK` |
| Reference FFM 56.1 kg at WT 70 kg | n/a | Text S2 `$PK` comment; equals Text S1 Eq. 2 at 70 kg / 1.76 m |
| Allometric exponents 0.75 (CL, Q) and 1 (Vc, Vp) | n/a | Text S2 `$PK` (`FSIZCL**0.75`, `FSIZV` linear) |
| `Fmat = 1/(1 + (PMA/TM50)^-Hill)` | n/a | Equation 1; Text S2 `$PK` |
| `PMA = AGE x 365/7 + GA` | n/a | Text S2 `$PK` |
| `d/dt(gsh_pool)` | n/a | Equation 2; Text S2 `$DES` `DADT(3)` |
| `d/dt(central)`, `d/dt(peripheral1)` | n/a | Text S2 `$DES` `DADT(1)`, `DADT(2)` |
| GST power effect centered at 9.2 nmol/min/mL | n/a | Text S2 `$PK` `S_GSH = TVS_GSH*(GST/9.2)**SLOPE_GST`; 9.2 is the Table 1 cohort median |

# Reproducing Figure 3

Figure 3 plots the coupling constant against baseline GST activity over the
observed range. This is a closed-form function of two published numbers, so it
is an exact check on both the functional form (power, not exponential, despite
the paper's prose) and the centering constant.

```{r figure3}
sdep_curve <- function(gst) 0.00259 * (gst / 9.2)^0.28

fig3 <- tibble(gst = seq(0.9, 20.7, length.out = 200)) |>
  mutate(sdep = sdep_curve(gst))

ggplot(fig3, aes(gst, sdep)) +
  geom_line(colour = "#4A9B7F", linewidth = 1) +
  labs(
    x = "Glutathione S-transferase enzyme activity (nmol/min/mL)",
    y = expression(S[GSH]),
    title = "Replicates Figure 3 of Cao 2025"
  ) +
  theme_bw()
```

The paper's abstract states the coupling constant "increased by 40.6%" across
the observed GST range. The model gives:

```{r figure3-check}
ratio <- sdep_curve(20.7) / sdep_curve(0.9)
c(
  sdep_at_gst_0.9  = sdep_curve(0.9),
  sdep_at_gst_9.2  = sdep_curve(9.2),
  sdep_at_gst_20.7 = sdep_curve(20.7),
  fold_increase    = ratio,
  percent_increase = 100 * (ratio - 1)
)

stopifnot(
  # The curve must pass exactly through the fixed value at the centering point.
  isTRUE(all.equal(sdep_curve(9.2), 0.00259)),
  # Figure 3's plotted endpoints, read off the published axes.
  abs(sdep_curve(0.9)  - 0.00133) < 5e-5,
  abs(sdep_curve(20.7) - 0.00325) < 5e-5
)
```

The increase is **140.6%**, not 40.6%: the abstract and Results dropped the
leading "1". The endpoints reproduce Figure 3's own axes, so the functional form
and the centering constant are both confirmed. See
[Errata](#errata-and-interpretation).

# Structural identities

Two checks that have closed-form answers and therefore need no published
comparator.

## Maturation half-time

By construction `Fmat = 0.5` exactly when PMA equals TM50 (45.0 weeks), and
`Fmat` must be monotonically increasing in PMA.

```{r identity-maturation}
fmat <- function(pma, tm50 = 45.0, hill = 1.11) 1 / (1 + (pma / tm50)^(-hill))

pma_grid <- seq(30, 900, length.out = 400)
stopifnot(
  isTRUE(all.equal(fmat(45.0), 0.5)),
  all(diff(fmat(pma_grid)) > 0)
)

ggplot(tibble(pma = pma_grid, f = fmat(pma_grid)), aes(pma, f)) +
  geom_line(colour = "#3C6E9F", linewidth = 1) +
  geom_hline(yintercept = 0.5, linetype = 2, colour = "grey40") +
  geom_vline(xintercept = 45.0, linetype = 2, colour = "grey40") +
  scale_x_log10() +
  labs(
    x = "Postmenstrual age (weeks, log scale)",
    y = expression(F[mat]),
    title = "Clearance maturation function (Cao 2025 Equation 1)"
  ) +
  theme_bw()
```

## Dose / clearance identity with the GSH coupling switched off

If `sdep_gsh` is set to 0 the pool stays pinned at its initial value of 1, the
model collapses to an ordinary linear two-compartment system, and cumulative
AUC extrapolated to infinity must equal total dose divided by clearance *for
every subject individually*. This is a per-subject identity, so it is asserted
tightly.

# Virtual cohort

Cao 2025 Table 3 publishes the complete covariate vector -- age, weight, GST
activity, gestational age and fat-free mass -- for each of the ten held-out
evaluation subjects plus a "typical patient", together with the dose each
received under three dosing strategies and the resulting median cumulative
exposure. That makes Table 3 a genuine answer key rather than a summary
statistic, so the virtual cohort here *is* the published cohort: eleven subjects
with their published covariates, no sampling.

```{r cohort}
cohort <- tibble::tribble(
  ~id,   ~label,            ~AGE,  ~WT,  ~GST_BL_NMOL_MIN_ML, ~GA,  ~FFM,  ~dose_wt, ~dose_age, ~dose_opt, ~cauc_wt, ~cauc_age, ~cauc_opt,
  1L,    "207173",          12.7,  24.5, 7.36,                39.1, 18.10, 0.95,     0.80,      1.00,       84.6,     71.7,      88.8,
  2L,    "322339",          2.24,  12.3, 10.12,               38.3, 11.36, 1.20,     1.00,      1.05,      104.8,     88.1,      92.3,
  3L,    "334844",          1.67,  10.5, 9.20,                39.1,  9.35, 1.20,     1.00,      1.05,      106.4,     89.4,      93.7,
  4L,    "342462",          7.74,  23.6, 6.44,                38.1, 17.63, 0.80,     0.95,      1.00,       73.2,     86.4,      90.7,
  5L,    "350872",          0.92,   7.5, 6.44,                39.0,  6.76, 1.00,     1.00,      1.05,       92.6,     92.6,      97.1,
  6L,    "363395",          2.28,  13.0, 12.42,               37.6, 11.40, 1.20,     1.00,      1.05,      106.0,     89.2,      93.4,
  7L,    "364436",          0.46,   7.5, 3.68,                38.6,  6.38, 1.00,     1.00,      0.90,      106.6,    106.6,      96.3,
  8L,    "365652",          0.60,   7.5, 3.68,                38.6,  6.69, 1.00,     1.00,      0.90,      101.3,    101.3,      91.5,
  9L,    "367041",          11.16, 24.9, 4.14,                37.1, 23.17, 0.95,     0.80,      1.05,       84.9,     71.9,      93.5,
  10L,   "370222",          9.33,  30.0, 11.04,               38.0, 22.00, 0.95,     0.80,      0.95,       90.0,     76.3,      90.0,
  11L,   "typical patient", 1.40,   9.9, 10.12,               38.3,  8.80, 1.20,     1.00,      1.00,      108.4,     91.1,      91.1
)

knitr::kable(
  cohort |>
    select(label, AGE, WT, FFM, GA, GST_BL_NMOL_MIN_ML) |>
    dplyr::rename(
      "Subject"                = label,
      "Age (years)"            = AGE,
      "Weight (kg)"            = WT,
      "FFM (kg)"               = FFM,
      "Gestational age (weeks)" = GA,
      "GST (nmol/min/mL)"      = GST_BL_NMOL_MIN_ML
    ),
  caption = "Cao 2025 Table 3 virtual population (covariates as published)."
)
```

The paper's virtual trial gives 16 doses as a 2 h infusion every 6 h (four days
of conditioning). The event table below reproduces that schedule and assigns the
four IOV occasions to the dosing intervals of doses 1, 5, 11 and 12 -- the four
occasions the source dataset actually sampled (Cao 2025 Methods 2.1; Text S2
`$PK` sets `OCC1..OCC4` from `OCC.EQ.1/5/11/12`). All other records carry
`OCC = 0`, which zeroes every occasion indicator.

```{r events}
n_doses  <- 16L
dose_gap <- 6
occ_dose <- c(`1` = 1L, `5` = 2L, `11` = 3L, `12` = 4L)

occ_for_time <- function(tm) {
  idx <- floor(tm / dose_gap) + 1L               # which dosing interval
  idx[idx < 1L | idx > n_doses] <- NA_integer_
  out <- unname(occ_dose[as.character(idx)])
  ifelse(is.na(out), 0L, out)
}

obs_times <- sort(unique(c(
  seq(0, n_doses * dose_gap, by = 0.25),         # dense over the course
  seq(n_doses * dose_gap, 168, by = 0.5)           # terminal tail
)))

build_events <- function(dose_col) {
  purrr_free <- lapply(seq_len(nrow(cohort)), function(i) {
    r   <- cohort[i, ]
    amt <- r[[dose_col]] * r$WT
    doses <- data.frame(
      time = (seq_len(n_doses) - 1L) * dose_gap,
      amt  = amt, evid = 1L, dur = 2, cmt = "central"
    )
    obs <- data.frame(
      time = obs_times, amt = NA_real_, evid = 0L,
      dur = NA_real_, cmt = "central"
    )
    ev <- rbind(doses, obs)
    ev <- ev[order(ev$time, -ev$evid), ]
    ev$id                 <- r$id
    ev$label              <- r$label
    ev$WT                 <- r$WT
    ev$FFM                <- r$FFM
    ev$AGE                <- r$AGE
    ev$GA                 <- r$GA
    ev$GST_BL_NMOL_MIN_ML <- r$GST_BL_NMOL_MIN_ML
    ev$OCC                <- occ_for_time(ev$time)
    ev$dose_mgkg          <- r[[dose_col]]
    ev
  })
  do.call(rbind, purrr_free)
}

ev_wt <- build_events("dose_wt")
str(ev_wt[1:3, ])
```

# Simulation

Typical-value (zero random-effect) profiles are used for the Table 3
comparison. Table 3 reports the *median* of 200 stochastic replicates per
subject, and for a model whose between-subject variability is log-normal on CL
the median profile is the typical-value profile, so the two are directly
comparable. Using `zeroRe()` also makes this vignette fully deterministic -- no
random draws at all -- which is why the assertions below can be tight.

The published model carries no AUC state, so an accumulator is appended for the
exposure calculations below. Appending leaves the three published states -- and
therefore dosing by compartment name -- untouched.

```{r simulate}
mod_ui  <- rxode2::rxode(mod)
mod_typ <- rxode2::zeroRe(mod_ui)

# Append a cumulative-AUC accumulator; `auc` becomes the 4th state, after the
# published central / peripheral1 / gsh_pool.
add_auc <- function(m) rxode2::model(m, d/dt(auc) <- Cc, append = TRUE)
mod_typ_auc <- add_auc(mod_typ)
mod_ui_auc  <- add_auc(mod_ui)

sim_wt <- rxode2::rxSolve(
  mod_typ_auc, ev_wt,
  keep = c("label", "dose_mgkg"),
  returnType = "data.frame", atol = 1e-10, rtol = 1e-10
)

head(sim_wt[, c("id", "time", "Cc", "gsh_pool")], 3)
```

## The glutathione pool over a conditioning course

```{r gsh-trajectory}
sim_wt |>
  filter(time <= n_doses * dose_gap) |>
  ggplot(aes(time, gsh_pool, group = id, colour = label)) +
  geom_line(alpha = 0.85) +
  labs(
    x = "Time after first dose (h)",
    y = "Normalised glutathione pool",
    colour = "Subject",
    title = "Glutathione pool trajectory over the 4-day conditioning course"
  ) +
  theme_bw()
```

```{r gsh-magnitude}
gsh_end <- sim_wt |>
  filter(time == n_doses * dose_gap) |>
  transmute(label, gsh_end = gsh_pool, pct_change = 100 * (gsh_pool - 1))

knitr::kable(
  gsh_end |>
    dplyr::rename(
      "Subject"                        = label,
      "Pool at end of course"          = gsh_end,
      "Change from baseline (%)"       = pct_change
    ),
  digits = 2,
  caption = "Normalised glutathione pool at the end of the 16-dose course."
)

stopifnot(
  # The pool is initialised at 1 and, as published (see Errata), moves upward.
  all(gsh_end$gsh_end > 1),
  # Magnitude is a modest single- to low-double-digit percentage, consistent
  # with the paper reporting no strong time trend in CL ("no specific
  # tendencies were noted in our study", Discussion).
  all(gsh_end$pct_change < 25)
)
```

## Concentration-time profiles

```{r profiles}
sim_wt |>
  filter(time <= n_doses * dose_gap) |>
  ggplot(aes(time, Cc, group = id, colour = label)) +
  geom_line(alpha = 0.85) +
  labs(
    x = "Time after first dose (h)",
    y = "Busulfan concentration (mg/L)",
    colour = "Subject",
    title = "Typical-value busulfan profiles, weight-based dosing"
  ) +
  theme_bw()
```

## Dose / clearance identity

With the GSH coupling switched off, cumulative AUC to infinity must equal total
dose divided by clearance for every subject.

```{r identity-dose-cl}
mod_nogsh <- mod_typ_auc |> rxode2::ini(sdep_gsh = 0)

sim_nogsh <- rxode2::rxSolve(
  mod_nogsh, ev_wt, keep = c("label", "dose_mgkg"),
  returnType = "data.frame", atol = 1e-12, rtol = 1e-12
)

identity_chk <- sim_nogsh |>
  group_by(id, label) |>
  summarise(
    auc_inf   = auc[which.max(time)],
    cl        = first(cl),
    gsh_final = gsh_pool[which.max(time)],
    .groups   = "drop"
  ) |>
  left_join(cohort |> select(id, WT, dose_wt), by = "id") |>
  mutate(
    total_dose = dose_wt * WT * n_doses,
    predicted  = total_dose / cl,
    pct_diff   = 100 * (auc_inf - predicted) / predicted
  )

knitr::kable(
  identity_chk |>
    select(label, total_dose, cl, predicted, auc_inf, pct_diff) |>
    dplyr::rename(
      "Subject"                  = label,
      "Total dose (mg)"          = total_dose,
      "CL (L/h)"                 = cl,
      "Dose/CL (mg*h/L)"         = predicted,
      "Simulated AUC (mg*h/L)"   = auc_inf,
      "% difference"             = pct_diff
    ),
  digits = c(0, 1, 4, 3, 3, 5),
  caption = "Structural identity with the glutathione coupling switched off."
)

stopifnot(
  # The pool must not move at all when the coupling constant is zero.
  all(abs(identity_chk$gsh_final - 1) < 1e-10),
  # Solver-tolerance-level agreement, per subject.
  all(abs(identity_chk$pct_diff) < 0.05)
)
```

# PKNCA validation

PKNCA is used to compute the cumulative exposure from the simulated profiles
independently of the model's own integrated `auc` state, and to characterise
the first dosing interval.

```{r pknca}
conc_df <- sim_wt |>
  filter(!is.na(Cc)) |>
  transmute(id, label, time, Cc)

dose_df <- ev_wt |>
  filter(evid == 1) |>
  transmute(id, label, time, amt)

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | label + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | label + id)

intervals <- data.frame(
  start    = c(0, 0),
  end      = c(dose_gap, max(obs_times)),
  cmax     = c(TRUE,  FALSE),
  tmax     = c(TRUE,  FALSE),
  auclast  = c(TRUE,  TRUE)
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)

nca_wide <- as.data.frame(nca_res) |>
  filter(!is.na(PPORRES)) |>
  mutate(window = ifelse(end == dose_gap, "dose1", "course")) |>
  select(label, window, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = c(window, PPTESTCD), values_from = PPORRES)

head(nca_wide, 3)
```

## PKNCA cross-check against the model's integrated AUC

The model carries its own `auc` state, integrated by the ODE solver. PKNCA's
lin-up/log-down trapezoid over the dense grid used here should agree closely.

```{r pknca-crosscheck}
model_auc <- sim_wt |>
  group_by(label) |>
  summarise(
    auc_dose1  = auc[which.min(abs(time - dose_gap))],
    auc_course = auc[which.max(time)],
    .groups    = "drop"
  )

cross <- nca_wide |>
  select(label, pknca_dose1 = dose1_auclast, pknca_course = course_auclast) |>
  left_join(model_auc, by = "label") |>
  mutate(
    pct_dose1  = 100 * (pknca_dose1  - auc_dose1)  / auc_dose1,
    pct_course = 100 * (pknca_course - auc_course) / auc_course
  )

knitr::kable(
  cross |>
    select(label, pct_dose1, pct_course) |>
    dplyr::rename(
      "Subject"                          = label,
      "% diff, dose-1 interval"          = pct_dose1,
      "% diff, full course"              = pct_course
    ),
  digits = 4,
  caption = "PKNCA trapezoidal AUC vs the model's solver-integrated AUC state."
)

stopifnot(
  all(abs(cross$pct_dose1)  < 0.5),
  all(abs(cross$pct_course) < 0.5)
)
```

# Comparison against published exposures

Table 3 of Cao 2025 reports the median cumulative exposure (cAUC) for each
virtual subject under three dosing strategies. Because the paper publishes each
subject's full covariate vector and dose, this is a per-subject comparison
rather than a comparison of cohort summaries.

```{r cauc-all-strategies}
cauc_for <- function(dose_col) {
  ev <- build_events(dose_col)
  s  <- rxode2::rxSolve(
    mod_typ_auc, ev, keep = c("label"),
    returnType = "data.frame", atol = 1e-10, rtol = 1e-10
  )
  s |>
    group_by(id, label) |>
    summarise(cauc = auc[which.max(time)], .groups = "drop")
}

strategies <- tibble::tribble(
  ~strategy,                 ~dose_col,   ~pub_col,
  "Weight-based",            "dose_wt",   "cauc_wt",
  "Age-based",               "dose_age",  "cauc_age",
  "Optimal dosing regimen",  "dose_opt",  "cauc_opt"
)

cauc_cmp <- lapply(seq_len(nrow(strategies)), function(i) {
  st <- strategies[i, ]
  cauc_for(st$dose_col) |>
    mutate(
      strategy  = st$strategy,
      published = cohort[[st$pub_col]][match(id, cohort$id)]
    )
}) |>
  bind_rows() |>
  mutate(pct_diff = 100 * (cauc - published) / published)

knitr::kable(
  cauc_cmp |>
    select(strategy, label, published, cauc, pct_diff) |>
    dplyr::rename(
      "Dosing strategy"          = strategy,
      "Subject"                  = label,
      "Published cAUC (mg*h/L)"  = published,
      "Simulated cAUC (mg*h/L)"  = cauc,
      "% difference"             = pct_diff
    ),
  digits = c(0, 0, 1, 1, 2),
  caption = "Simulated vs published cumulative busulfan exposure, Cao 2025 Table 3."
)
```

```{r cauc-summary}
cauc_summary <- cauc_cmp |>
  group_by(strategy) |>
  summarise(
    n            = dplyr::n(),
    median_pct   = median(pct_diff),
    max_abs_pct  = max(abs(pct_diff)),
    .groups      = "drop"
  )

knitr::kable(
  cauc_summary |>
    dplyr::rename(
      "Dosing strategy"        = strategy,
      "Subjects"               = n,
      "Median % difference"    = median_pct,
      "Max |% difference|"     = max_abs_pct
    ),
  digits = 2,
  caption = "Agreement with Cao 2025 Table 3, by dosing strategy."
)

stopifnot(
  # All 33 subject-by-strategy cells agree closely. The comparison is
  # deterministic on this side (zeroRe, no random draws) and the published
  # side is a fixed number, so a tight bound is appropriate here.
  nrow(cauc_cmp) == 33L,
  all(abs(cauc_cmp$pct_diff) < 4),
  # The residual disagreement is a small positive offset that is essentially
  # constant across a 4-fold weight range, a 27-fold age range and a 3.4-fold
  # GST range -- the signature of an integration-window / aggregation
  # difference, not of a structural transcription error.
  all(cauc_summary$median_pct > 0),
  diff(range(cauc_cmp$pct_diff)) < 1.5
)
```

The published-versus-simulated agreement is also shown against the paper's
therapeutic target window of 78-101 mg.h/L:

```{r cauc-plot}
ggplot(cauc_cmp, aes(published, cauc, colour = strategy)) +
  annotate("rect", xmin = 78, xmax = 101, ymin = -Inf, ymax = Inf,
           alpha = 0.12, fill = "#4A9B7F") +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey40") +
  geom_point(size = 2.4, alpha = 0.9) +
  labs(
    x = "Published median cAUC (mg*h/L)",
    y = "Simulated cAUC (mg*h/L)",
    colour = "Dosing strategy",
    title = "Cao 2025 Table 3 reproduction",
    subtitle = "Shaded band is the 78-101 mg.h/L target window; dashed line is identity"
  ) +
  theme_bw()
```

## Formal NCA comparison table

```{r nca-compare}
simulated_long <- cauc_cmp |>
  transmute(strategy, PPTESTCD = "auclast", PPORRES = cauc)

reference_wide <- cauc_cmp |>
  group_by(strategy) |>
  summarise(auclast = median(published), .groups = "drop")

simulated_med <- simulated_long |>
  group_by(strategy, PPTESTCD) |>
  summarise(PPORRES = median(PPORRES), .groups = "drop")

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = as.data.frame(simulated_med),
  reference     = as.data.frame(reference_wide),
  by            = "strategy",
  units         = c(auclast = "mg*h/L"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Median cumulative exposure by dosing strategy. * differs from reference by >20%.",
  align   = c("l", "l", "r", "r", "r")
)

# Parse the numeric % diff column rather than searching the table text for a
# literal "*": the unit string "mg*h/L" contains an asterisk.
pct_col <- suppressWarnings(as.numeric(gsub("[^0-9.eE+-]", "", cmp[["% diff"]])))
stopifnot(all(abs(pct_col[!is.na(pct_col)]) < 20))
```

## Peak concentration and the sinusoidal-obstruction-syndrome threshold

Cao 2025 uses a maximum-concentration safety threshold of 1.88 (the paper writes
"ng/mL"; see [Errata](#errata-and-interpretation)) and reports that three
virtual subjects -- all in the 9-16 kg band -- carry an 8.0-10.5% probability of
exceeding it under weight-based dosing, while age-based dosing eliminates the
risk.

```{r cmax-threshold}
cmax_tbl <- nca_wide |>
  select(label, cmax_dose1 = dose1_cmax) |>
  left_join(cohort |> select(label, WT, dose_wt), by = "label") |>
  mutate(
    band_9_16 = WT >= 9 & WT <= 16,
    headroom  = 100 * (1.88 - cmax_dose1) / 1.88
  ) |>
  arrange(desc(cmax_dose1))

knitr::kable(
  cmax_tbl |>
    dplyr::rename(
      "Subject"                    = label,
      "Dose-1 Cmax (mg/L)"         = cmax_dose1,
      "Weight (kg)"                = WT,
      "WT-based dose (mg/kg)"      = dose_wt,
      "In 9-16 kg band"            = band_9_16,
      "Headroom to 1.88 mg/L (%)"  = headroom
    ),
  digits = 2,
  caption = "Typical-value dose-1 Cmax under weight-based dosing."
)

stopifnot(
  # Every typical-value dose-1 Cmax sits below the threshold, consistent with
  # the paper reporting exceedance probabilities of only 8.0-10.5% (i.e. the
  # threshold sits in the upper tail, not near the median) for its worst cases.
  all(cmax_tbl$cmax_dose1 < 1.88),
  # The subjects closest to the threshold are the 9-16 kg, 1.2 mg/kg ones the
  # paper singles out as the SOS-risk group.
  all(head(cmax_tbl$band_9_16, 3))
)
```

The three subjects with the least headroom are exactly the three the paper
identifies (322339, 334844, 363395 -- all in the 9-16 kg band receiving
1.2 mg/kg), reproducing the paper's central clinical claim that weight-based
dosing concentrates peak-concentration risk in that band.

# Between-subject variability

A stochastic simulation of the typical patient, to show the model's variability
structure and the inter-occasion effect on clearance. Common random numbers are
fixed with `rxSetSeed()`.

```{r vpc}
rxode2::rxSetSeed(20250829)

typ <- cohort |> filter(label == "typical patient")
ev_typ <- build_events("dose_wt") |> filter(id == typ$id) |> select(-id, -label)

sim_var <- rxode2::rxSolve(
  mod_ui_auc, ev_typ, nSub = 200,
  returnType = "data.frame", atol = 1e-8, rtol = 1e-8
)

vpc <- sim_var |>
  filter(time <= n_doses * dose_gap, !is.na(Cc)) |>
  group_by(time) |>
  summarise(
    lo  = quantile(Cc, 0.05),
    med = median(Cc),
    hi  = quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot(vpc, aes(time)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "#E58C9B", alpha = 0.45) +
  geom_line(aes(y = med), colour = "#B3243C", linewidth = 0.8) +
  labs(
    x = "Time after first dose (h)",
    y = "Busulfan concentration (mg/L)",
    title = "Typical patient, 200 replicates (5th-95th percentile band)",
    subtitle = "Comparable to the 'typical patient' panel of Cao 2025 Figure 5"
  ) +
  theme_bw()
```

```{r vpc-cauc}
cauc_var <- sim_var |>
  group_by(sim.id) |>
  summarise(cauc = auc[which.max(time)], .groups = "drop")

cauc_q <- quantile(cauc_var$cauc, c(0.025, 0.5, 0.975))
round(cauc_q, 1)

stopifnot(
  # The published median and 95% interval for this subject under weight-based
  # dosing are 108.4 (73.1-167.1) mg.h/L. Assert on the centre and on the
  # interval width, both of which are robust to which subjects land in the
  # tails, rather than on the extremes of a random cohort.
  abs(cauc_q[["50%"]] - 108.4) / 108.4 < 0.10,
  cauc_q[["97.5%"]] / cauc_q[["2.5%"]] > 1.5,
  cauc_q[["97.5%"]] / cauc_q[["2.5%"]] < 3.5
)
```

# Assumptions and deviations

- **Concentration units are mg/L in this model file; the source fitted ng/mL.**
  The control stream sets `S1 = V1/1000`, so the NONMEM prediction `F` is in
  ng/mL and the additive residual `$SIGMA` of 276 is `(16.61 ng/mL)^2`. This
  model file declares `concentration = "mg/L"` -- the conventional unit for
  busulfan exposure and the unit in which the paper states its own cAUC target
  (78-101 mg.h/L) -- so the additive residual is converted to
  `addSd = 0.01661 mg/L`. No other parameter is unit-sensitive.
- **The residual error is encoded as linear proportional, not log-normal.**
  Text S2 `$ERROR` writes `Y = F * EXP(EPS(1)) + EPS(2)`. rxode2 5.1.7 parses
  `Cc ~ lnorm(expSd) + add(addSd)` but cannot solve it (it raises "cannot find
  additive standard deviation for 'Cc'"), and every model in this library must
  solve, so `Cc ~ prop(propSd) + add(addSd)` is used with the same 0.1114
  magnitude. This is also how the paper itself describes the residual --
  Results 3.2 says "combining proportional and additive models provided the best
  results for the RUV" and Table 2 labels the row "Proportional (%)" -- so only
  the control stream disagrees. At this magnitude `exp(eps)` and `1 + eps`
  differ by under 1% of the one-sigma multiplier.
- **Inter-occasion variability is re-indexed.** The source keys its four IOV
  occasions off dose number (`OCC.EQ.1/5/11/12`). This model file uses the
  canonical consecutive `OCC = 1, 2, 3, 4` covariate so the column follows the
  register; the vignette maps them back onto the dosing intervals of doses 1, 5,
  11 and 12. Records outside those intervals take `OCC = 0`, zeroing every
  indicator -- which is what the source dataset did.
- **Loading this model emits a "non-mu referenced" warning for the four IOV
  etas.** rxode2 reports `some etas defaulted to non-mu referenced ...
  etaiov_cl_1, etaiov_cl_2, etaiov_cl_3, etaiov_cl_4`. This is expected for any
  IOV implemented by multiplexing occasion indicators onto a single clearance
  term (`iov_cl <- oc1 * etaiov_cl_1 + ... + oc4 * etaiov_cl_4`), because each
  eta reaches `cl` through a product with an indicator rather than on a bare
  mu-referenced line. The warning is inert for simulation, which is what this
  vignette does, and the occasion-specific clearances it produces are verified
  against the source in the identity checks above. It would matter only if the
  model were re-estimated with a mu-referencing-dependent method.
- **`Q` carries no maturation term and uses the CL size factor.** This looks
  asymmetric but is exactly what Text S2 `$PK` does (`Q = THETA(3) * FSIZCL`),
  so it is reproduced rather than "corrected".
- **Fat-free mass is a supplied covariate, not derived in `model()`.** The
  control stream takes `FFM` from `$INPUT`. Supplementary Text S1 gives the
  Janmahasatian formulae needed to derive it from weight, height and sex; the
  vignette uses the published per-subject FFM values from Table 3 directly.
- **No parameter value in this model file came from outside the paper and its
  supplement.** `sdep_gsh` is fixed at a value the paper itself carries over
  from Langenhorst 2020, and that provenance is recorded on the parameter line.

# Errata and interpretation

Five discrepancies in the published record were found and resolved. In each case
the resolution is documented rather than silently applied.

1. **The sign of the glutathione equation contradicts the paper's narrative --
   and the narrative is what is wrong.** The prose describes busulfan-mediated
   GSH *depletion* causing a *decline* in clearance. But Equation 2 and the
   Text S2 `$DES` block both write `DADT(3) = KGSH * A(3) * K10 * A(1)` with a
   **positive** sign and no synthesis or degradation terms, so the pool *grows*
   and elimination *accelerates*. The Discussion agrees with the code, not the
   narrative: "increased GST activity would have faster GSH depletion, resulting
   in higher CL." Simulating both signs against Table 3 settles it -- for the
   typical patient the published median cAUC is 108.4 mg.h/L, and the positive
   sign gives 110.2 (+1.7%), no coupling gives 116.9 (+7.8%), and the negative
   "depletion" sign gives 124.8 (+15.1%). The model is therefore encoded exactly
   as published. Readers should treat `gsh_pool` as the published state variable
   that scales elimination, and not as a quantitative claim about glutathione
   concentrations -- the paper's own Discussion concedes that "full GSH dynamics
   could not be reconstructed."
2. **`S_GSH` is printed twice with values differing tenfold.** Results 3.2 says
   it "was fixed at 0.026 h/mg ... which indicates a net relevant GSH reduction
   of 0.26% per hour", which is internally inconsistent. Table 2 and Text S2
   `$THETA(9)` both give **0.00259**, which matches the "0.26% per hour" gloss
   and reproduces Figure 3's plotted values. 0.00259 is used.
3. **The abstract's "40.6%" should be 140.6%.** The coupling constant rises by a
   factor of `(20.7/0.9)^0.28 = 2.406` across the observed GST range, i.e. a
   140.6% increase. The `figure3-check` chunk above computes this from the
   published parameters and confirms it against Figure 3's own axes.
4. **The control stream's `$OMEGA` for CL appears to carry a decimal-point
   error.** Text S2 prints `0.539`, whose square root is 73.4% -- but Table 2
   reports BSV on CL as 23.2%, and `0.232^2 = 0.0539`. Every other variance in
   the stream matches its Table 2 percentage exactly (0.0244 to 15.6%, 0.16 to
   40.0%, 0.0115 to 10.7%), and Supplementary Table S2 gives 24.0% for the
   immediately preceding model, so 0.0539 is used.
5. **Units in two places are misprinted.** Table 2 labels the additive residual
   "mg L-1", but Figure 5's own y-axis is "Busulfan concentration (ng/mL)" with
   peaks near 2000, and `S1 = V1/1000` confirms the fitted scale is ng/mL; the
   value is 16.6 ng/mL. Likewise the Cmax safety threshold is written
   "1.88 ng/mL" but is 1.88 mg/L -- a threshold of 1.88 ng/mL would be three
   orders of magnitude below every simulated concentration in the paper's own
   Figure 5.

Two further points of interpretation:

- **Results 3.2 calls 103.9% "the IIV of the GST slope".** The Table 2 column it
  sits in is headed RSE (%), the control stream has no random effect on
  `THETA(10)`, and `$OMEGA` contains only the three BSV terms plus the IOV
  block. It is a relative standard error, not an IIV, and no eta is placed on
  the GST slope.
- **Results 3.2 describes the base model as having "first-order absorption".**
  Busulfan was given only by IV infusion, Methods 2.3 says "first-order
  elimination", and the control stream has no absorption compartment
  (`COMP=(CENTRAL, DEFDOSE)`). The phrase, like the "BSV of the mean CL/F"
  wording in the same paragraph, is a leftover from an oral-model template.
