---
title: "Radiation + radiosensitizer tumor model (Cardilin 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Radiation + radiosensitizer tumor model (Cardilin 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
set.seed(1234)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Cardilin_2018_radiation_radiosensitizer_mouse"))$reference`
* Description: `r rxode2::rxode(readModelDb("Cardilin_2018_radiation_radiosensitizer_mouse"))$description`
* Article (open access): <https://doi.org/10.1002/psp4.12268>
* Supplement (Europe PMC, PMC5784742): <https://www.ebi.ac.uk/europepmc/webservices/rest/PMC5784742/supplementaryFiles>

This is a preclinical tumor-growth-inhibition (TGI) model for combination
therapy with ionizing radiation and a radiosensitizing compound (anonymized
"RS1" in the source), fit to FaDu head-and-neck xenograft data in mice. The
proliferating compartment grows logistically and feeds a three-stage
natural-death transit chain. Radiation acts through the linear-quadratic (LQ)
surviving-fraction theory as a near-instantaneous transfer of proliferating
cells into an irradiated chain (which is allowed at most one further division
before dying). The radiosensitizer's one-compartment plasma PK drives both a
small enhancement of natural cell death and a strong synergistic enhancement of
the radiation kill.

# Population

The structural model was estimated from a FaDu (human head-and-neck
squamous-cell-carcinoma) xenograft experiment in female CD1 nu/nu or NMRI nu/nu
mice (Supplementary Material B of the source). The primary tumor dataset
comprised 80 mice in eight groups of ten: (A) vehicle, (B) radiation 2 Gy,
(C-E) RS1 monotherapy 10/50/200 mg/kg, and (F-H) combination of 2 Gy + RS1
10/50/200 mg/kg, dosed once daily on days 3-7. RS1 was given orally 10 minutes
before each irradiation. A second 10-mouse dataset (2 Gy + RS1 25 mg/kg, five
days per week for six weeks) was used for prediction (Figure 5 of the source).
The radiosensitizer PK was estimated separately from 147 tumor-bearing female
mice across five single-dose studies (Supplementary Table SI). Between-subject
variability was estimated for the initial tumor volume, the tumor capacity, and
the two PK parameters.

The same information is available programmatically via
`readModelDb("Cardilin_2018_radiation_radiosensitizer_mouse")$population`.

# Source trace

Per-parameter origins are recorded as in-file comments next to each `ini()`
entry. The table collects them for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lkg` (growth rate kg) | 0.50 1/day | Table 1 |
| `lkk` (kill rate kk) | 0.28 1/day | Table 1 |
| `lcap` (capacity K) | 2200 mm^3 | Table 1 |
| `lv0` (initial volume V0) | 40.0 mm^3 | Table 1 |
| `lalpha` (LQ alpha) | 0.08 1/Gy | Table 1 |
| `abratio` (alpha/beta, fixed) | 10 Gy | Methods + Table 1 footnote |
| `laDeath` (natural-death stim., paper "a") | 0.09 mL/ug | Table 1 + Discussion ("9% per ug/mL") |
| `lbRad` (radiation synergy, paper "b") | 0.45 mL/ug | Table 1 + Discussion ("45% per ug/mL") |
| `lke` (RS1 elimination) | 0.21 1/h (x 24 = 5.04 1/day) | Supplementary Table SI |
| `lvf` (RS1 apparent V/F) | 9.51 L/kg | Supplementary Table SI |
| `radDose` (per fraction) | 2 Gy | Methods |
| `etalv0` / `etalcap` | BSV 21% / 32% | Table 1 |
| `etalvf` / `etalke` | BSV 88% / 15% | Supplementary Table SI |
| `propSd` | 23% | Table 1 (sigma) |
| ODE system (Eq. 6), ICs (Eq. 7) | n/a | Methods + Mathematica supplement (`s008`) |
| PK solution `C(t)=D F / V exp(-ke t)` | n/a | Supplementary Material A, Eq. 18 |

# Simulation set-up

Radiation is delivered as a unit bolus (`amt = 1`) into the `radDepot`
compartment at each irradiation time; the per-fraction radiation dose (Gy) is
the model parameter `radDose` (default 2). RS1 is dosed (mg/kg) into the
`central` compartment. The observation `tumor_vol` is the total tumor volume
(sum of all six tumor compartments, mm^3).

```{r model}
mod <- readModelDb("Cardilin_2018_radiation_radiosensitizer_mouse")
mod_typ <- rxode2::zeroRe(mod)   # typical-value (no between-subject variability)

# Build an event table for one treatment group on the days-3-7 schedule.
make_group <- function(rs_dose = 0, irradiate = FALSE, days = 3:7,
                       tmax = 30, by = 0.25) {
  ev <- rxode2::et(seq(0, tmax, by = by))
  if (rs_dose > 0)
    for (d in days) ev <- rxode2::et(ev, amt = rs_dose, cmt = "central",
                                     time = d - 10/1440)  # 10 min before IR
  if (irradiate)
    for (d in days) ev <- rxode2::et(ev, amt = 1, cmt = "radDepot", time = d)
  ev
}
```

# Replicate Figure 2 -- tumor volume by treatment group

```{r figure-2}
groups <- list(
  "A: vehicle"            = make_group(0,   FALSE),
  "B: radiation 2 Gy"     = make_group(0,   TRUE),
  "E: RS1 200 mg/kg"      = make_group(200, FALSE),
  "F: 2 Gy + RS1 10"      = make_group(10,  TRUE),
  "G: 2 Gy + RS1 50"      = make_group(50,  TRUE),
  "H: 2 Gy + RS1 200"     = make_group(200, TRUE)
)
traj <- bind_rows(lapply(names(groups), function(g) {
  s <- as.data.frame(rxode2::rxSolve(mod_typ, groups[[g]]))
  data.frame(time = s$time, tumor_vol = s$tumor_vol, group = g)
}))

ggplot(traj, aes(time, tumor_vol)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~group) +
  labs(x = "Time (day)", y = expression("Tumor volume (mm"^3*")"),
       title = "Replicates Figure 2 of Cardilin et al. (2018)",
       caption = "Typical-value tumor growth; radiation 2 Gy/dose on days 3-7.") +
  theme_bw()
```

Vehicle (A) grows approximately exponentially toward the capacity; RS1
monotherapy (E) deviates only slightly from vehicle; radiation alone (B) slows
growth considerably but does not produce regression; and combination therapy
shrinks the tumor in a dose-dependent way, with the highest dose (H) driving
rapid, near-complete eradication. This matches the source narrative and
Figure 2.

# Quantitative validation -- per-fraction kill fractions

The source reports that a single 2 Gy fraction kills 17% of proliferating
cells, rising to 24%, 46%, and 85% in combination with RS1 10, 50, and
200 mg/kg (Discussion). We reproduce these by measuring the drop in the
cycling-cells compartment across one fraction.

```{r kill-fractions}
kill_one <- function(rs_dose) {
  ev <- rxode2::et(seq(0, 4, by = 0.001))
  if (rs_dose > 0) ev <- rxode2::et(ev, amt = rs_dose, cmt = "central",
                                    time = 3 - 10/1440)
  ev <- rxode2::et(ev, amt = 1, cmt = "radDepot", time = 3)
  s <- as.data.frame(rxode2::rxSolve(mod_typ, ev, maxsteps = 1e6,
                                     atol = 1e-9, rtol = 1e-9))
  pre  <- s$cycling_cells[which.min(abs(s$time - 2.99))]
  post <- s$cycling_cells[which.min(abs(s$time - 3.05))]  # after the fast kill spike
  100 * (1 - post / pre)
}
kill_tab <- data.frame(
  scenario  = c("2 Gy alone", "2 Gy + RS1 10", "2 Gy + RS1 50", "2 Gy + RS1 200"),
  simulated = round(sapply(c(0, 10, 50, 200), kill_one), 0),
  published = c(17, 24, 46, 85)
)
knitr::kable(kill_tab, col.names = c("Scenario", "Simulated kill (%)",
                                     "Published kill (%)"),
             caption = "Per-fraction proliferating-cell kill: model vs source.")
```

The combination kill fractions reproduce the published values; the 2 Gy-alone
value reads slightly low because it is measured shortly after the kill, during
which the surviving cells simultaneously grow and die.

# Quantitative validation -- RS1 average plasma concentrations

The source reports average daily RS1 plasma concentrations of 0.52, 0.9, and
8.2 ug/mL for 25, 48, and 400 mg/kg (main text and Supplementary Material D).
We reproduce these from the one-compartment PK by averaging concentration over
a one-day dosing interval.

```{r conc-check}
cavg <- function(dose) {
  ev <- rxode2::et(amt = dose, cmt = "central", time = 0) |>
    rxode2::et(seq(0, 1, by = 0.001))
  s <- as.data.frame(rxode2::rxSolve(mod_typ, ev))
  # trapezoidal mean over [0, 1] day
  mean(0.5 * (head(s$Cc, -1) + tail(s$Cc, -1)))
}
conc_tab <- data.frame(
  dose_mgkg = c(25, 48, 400),
  simulated = round(sapply(c(25, 48, 400), cavg), 2),
  published = c(0.52, 0.9, 8.2)
)
knitr::kable(conc_tab, col.names = c("RS1 dose (mg/kg)",
                                     "Simulated mean conc (ug/mL)",
                                     "Published mean conc (ug/mL)"),
             caption = "Average daily RS1 plasma concentration: model vs source.")
```

# Replicate Figure 5 -- six-week combination prediction

The second (prediction) dataset used 2 Gy + RS1 25 mg/kg five days per week
(Mon-Fri) for six weeks. We simulate the population (5th/50th/95th percentiles)
on this schedule.

```{r figure-5}
treat_days <- as.vector(outer(1:5, 7 * (0:5), `+`))  # Mon-Fri for 6 weeks
treat_days <- sort(treat_days)

ev6 <- rxode2::et(seq(0, 45, by = 0.5))
for (d in treat_days) {
  ev6 <- rxode2::et(ev6, amt = 25, cmt = "central", time = d - 10/1440)
  ev6 <- rxode2::et(ev6, amt = 1,  cmt = "radDepot", time = d)
}

sim6 <- rxode2::rxSolve(mod, ev6, nSub = 200, maxsteps = 1e6) |>
  as.data.frame()

vpc6 <- sim6 |>
  group_by(time) |>
  summarise(Q05 = quantile(tumor_vol, 0.05, na.rm = TRUE),
            Q50 = quantile(tumor_vol, 0.50, na.rm = TRUE),
            Q95 = quantile(tumor_vol, 0.95, na.rm = TRUE), .groups = "drop")

ggplot(vpc6, aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time (day)", y = expression("Tumor volume (mm"^3*")"),
       title = "Replicates Figure 5 of Cardilin et al. (2018)",
       caption = "2 Gy + RS1 25 mg/kg, Mon-Fri for 6 weeks; 200 virtual mice.") +
  theme_bw()
```

Consistent with Figure 5 and the source narrative, the clinically relevant
six-week schedule drives the tumor into sustained regression (most simulated
animals are eradicated by the end of treatment).

# Dimensional analysis

The model mixes day-based rates, mm^3 volumes, ug/mL concentrations, and Gy
radiation doses. Each ODE term is dimensionally consistent with
`d/dt(state) = volume/time = mm^3/day`:

| Term | Units | Notes |
|------|-------|-------|
| `kg * cycling_cells * logi` | (1/day)(mm^3)(-) = mm^3/day | `logi` dimensionless |
| `(1 + aDeath*Cc) * kk * cycling_cells` | (-)(1/day)(mm^3) = mm^3/day | `aDeath*Cc` = (mL/ug)(ug/mL) dimensionless |
| `killHaz * cycling_cells` | (1/day)(mm^3) = mm^3/day | see below |
| `lethal` | (-)(1/Gy * Gy) = dimensionless | LQ lethal lesions, `alpha*radDose + beta*radDose^2` |
| `krad * radDepot` | (1/day)(-) integrates to 1 per fraction | numerical delta-trigger |
| `Cc = central / vf` | (mg/kg)/(L/kg) = mg/L = ug/mL | RS1 plasma concentration |
| `ke * central` | (1/day)(mg/kg) = (mg/kg)/day | RS1 elimination |

The radiation kill hazard `killHaz = lethal * krad * radDepot` integrates over a
fraction to `lethal` (since the unit `radDepot` bolus has integral 1 under the
`krad` decay), so the proliferating pool is multiplied by the LQ surviving
fraction `exp(-lethal) = exp(-(1 + bRad*Cc)(alpha*D + beta*D^2))`.

# Assumptions and deviations

- **Metadata correction (year and name).** The dispatch metadata named the model
  `Cardilin_2017_CPT_Pharmacometrics_amp_Sys` (a queue-generation artifact in
  which the journal name was placed in the drug slot, and the online-first year
  2017 was used). The article's formal citation is *CPT Pharmacometrics Syst
  Pharmacol* **2018**;7(1):51-58 (published online 14 December 2017); 2018 is
  the volume/citation year used by PubMed (PMID 29218836) and matches the
  on-disk source filename. The compound is anonymized as "RS1" in the source, so
  the descriptive slot `radiation_radiosensitizer` and the `_mouse` species
  suffix are used.
- **Coefficient placement (paper "a" vs "b").** Table 1 of the source and the
  code variables in the Mathematica supplement label the two radiosensitizer
  coefficients oppositely. The placement here (natural-death stimulation
  = 0.09 mL/ug, radiation synergy = 0.45 mL/ug) follows the paper's quantitative
  text ("9% increase in natural cell death" and "45% increase in fraction
  killed" per ug/mL) and is confirmed by the published combination kill
  fractions (24/46/85% for 2 Gy + RS1 10/50/200 mg/kg), reproduced above.
- **RS1 apparent volume units.** Supplementary Table SI prints `V = 0.0095` with
  the label "mL/kg". On a ng/mL concentration basis this value is
  0.009507 (mg/kg)/(ng/mL) = 9.51 L/kg (the printed label is a mis-transcription).
  This value reproduces the reported average plasma concentrations (above), so
  9.51 L/kg is used.
- **Radiation as a near-instantaneous transfer.** The source models irradiation
  as Dirac-delta mass transfers. This is implemented with a fast-decaying unit
  trigger (`radDepot`, rate `krad = 500` 1/day, integral 1 per fraction) so the
  proliferating pool is multiplied by the LQ surviving fraction. `krad` is a
  solver convenience, not a fitted parameter; increasing it sharpens the
  approximation.
- **Between-subject variability scale.** The source reports BSV as
  `sqrt(omega^2) x 100` (Table 1 / Table SI footnotes), so log-scale variances
  are `(BSV%/100)^2` (V0 0.0441, K 0.1024, V/F 0.7744, ke 0.0225).
- **Convention-lint warnings (justified, not errors).** The mechanistic tumor
  compartments (`cycling_cells`, `damaged_cells1-3`, `irrad1-2`) and the radiation trigger
  (`radDepot`) are not canonical PK compartment names, mirroring the existing
  `oncology_xenograft_simeoni_2004` TGI model. The fitted output is total tumor
  volume (`tumor_vol`, a non-PK output), while `Cc` is correctly reserved for the
  RS1 plasma concentration. The units `mg/kg` (RS1 dose) and `ug/mL`
  (concentration) are both mass-based and reconcile through the apparent volume
  `V/F` (L/kg); the lint's prefix-level dimensional check is a false positive.
- **Initial total volume.** With the Table 1 estimates the initial total volume
  is `V0 (1 + (kk/kg) + (kk/kg)^2 + (kk/kg)^3)` = 82 mm^3; the source reports
  ~87 mm^3, the small difference arising from rounding of kg and kk.
- **Observed data.** The original animal-level tumor measurements are not
  publicly available; the figures above are model simulations on the published
  dosing schedules.
