---
title: "Enflicoxib (Cendros 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Enflicoxib (Cendros 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Cendros_2025_enflicoxib"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Article: <https://doi.org/10.3389/fvets.2025.1645857>\n")
```

`r ui$description`

Enflicoxib is a COX-2 selective NSAID licensed for canine osteoarthritis with an
unusual **once-weekly** oral posology. The posology works because the parent is
a precursor of a much longer-lived active pyrazol metabolite: the parent's
terminal half-life is 1.4 days while the metabolite's is 13.8 days, so the
metabolite carries the exposure between weekly doses.

The structural model, its parameter estimates, the inter-individual variability
and the residual error were all estimated by **Cendros 2022** (Vet Med Sci
8:31-45, doi:10.1002/vms3.670) in young healthy Beagle dogs. **Cendros 2025**
reproduces that full parameter set in its Table 1 and *externally validates* it,
unchanged, against sparse plasma samples from 83 client-owned dogs of any breed
with naturally occurring osteoarthritis treated weekly for six months. No
parameter was re-estimated in the 2025 paper; the Bayesian step used
`$ESTIMATION MAXEVAL=0` with the `POSTHOC` subroutine.

# Population

```{r population, echo = FALSE}
pop <- ui$population
tibble::tibble(Field = names(pop),
               Value = vapply(pop, function(x) paste(as.character(x), collapse = "; "),
                              character(1))) |>
  knitr::kable(caption = "Population metadata carried in the model file.")
```

Eighty-three client-owned dogs from the active-treatment arm of a blinded,
randomised, placebo-controlled multicentre field study in Portugal and Hungary
received at least one dose of enflicoxib and form the PK dataset (Cendros 2025
Results, "Study population"). Forty-eight (57.8%) were male; the mean age was
8.7 +/- 3 years (range 2-16) and the mean body weight 27.0 +/- 15 kg (range
4.9-64.9). Half the cohort (50.6%) was purebred, with more than 25 breeds
represented. Each dog contributed at most two plasma samples, one around day 44
and one around day 189, giving 142 concentration records in total.

The allometric reference weight of 9.9 kg is *not* this cohort's median -- it is
the median weight of the healthy Beagle population in which the parameters were
estimated, and it is read directly off the `(WGT/9.9)` term written into every
Cendros 2025 Table 1 parameter equation.

# Source trace

Per-parameter provenance is recorded as an in-file comment beside each `ini()`
entry in `inst/modeldb/specificDrugs/Cendros_2025_enflicoxib.R`. Collected here
for review:

| Equation / parameter | Value | Source location |
|---|---|---|
| `lfdepot` (F) | 1 FIX | Table 1, enflicoxib, "Relative bioavailability (F)" |
| `ltlag` (Tlag) | 0.249 h | Table 1, enflicoxib, "Lag Time (Tlag)" |
| `lka` (ka) | 0.220 1/h | Table 1, enflicoxib, "Absorption rate (ka)" |
| `lvc` (V/F, theta4) | 6.59 L | Table 1, enflicoxib, `V/F = theta4*(WGT/9.9)^1.0` |
| `lvp` (V5/F, theta11) | 27.70 L | Table 1, enflicoxib, `V5/F = theta11*(WGT/9.9)^1.0` |
| `lq` (CL5/F, theta10) | 12.1 L/h | Table 1, enflicoxib, `CL5/F = theta10*(WGT/9.9)^0.75` |
| `lcl_met` (CL2/F, theta6) | 0.520 L/h | Table 1, enflicoxib Elimination; footnote (dagger) "apparent clearance of irreversible biotransformation to pyrazol metabolite" |
| `lcl_nonmet` (CL1/F, theta5) | 0.193 L/h | Table 1, enflicoxib Elimination; footnote (dagger) "apparent clearance to other elimination pathways" |
| `lvc_pyrazol` (VM/F, theta4) | 6.59 L | Table 1, pyrazol metabolite, `VM/F = theta4*(WGT/9.9)^1.0` -- the *same* theta4 as the parent V/F |
| `lvp_pyrazol` (VMP/F, theta9) | 29.3 L | Table 1, pyrazol metabolite, `VMP/F = theta9*(WGT/9.9)^1.0` |
| `lvp2_pyrazol` (VM2/F, theta13) | 13.4 L | Table 1, pyrazol metabolite, `VM2/F = theta13*(WGT/9.9)^1.0` |
| `lq_pyrazol` (CLMP/F, theta8) | 13.2 L/h | Table 1, pyrazol metabolite, `CLMP/F = theta8*(WGT/9.9)^0.75` |
| `lq2_pyrazol` (CLM2/F, theta12) | 0.131 L/h | Table 1, pyrazol metabolite, `CLM2/F = theta12*(WGT/9.9)^0.75` |
| `lcl_pyrazol` (CLM/F, theta7) | 0.111 L/h | Table 1, pyrazol metabolite Elimination |
| `e_wt_cl` | 0.75 FIX | Table 1 (every clearance row `(WGT/9.9)^0.75`); Discussion confirms fixed a priori |
| `e_wt_vc` | 1.0 FIX | Table 1 (every volume row `(WGT/9.9)^1.0`); Discussion confirms fixed a priori |
| Allometric equation | `P_POP = theta1*(WGT/median(WGT))^theta2` | Methods, "Established pop PK model in healthy Beagle dogs" |
| IIV form | `P_i = P_POP*exp(eta_i)`, magnitude as %CV | Methods, same section |
| `etalfdepot` / `etaltlag` / `etalka` / `etalvc` / `etalcl_met` / `etalcl_pyrazol` | 39 / 68 / 101 / 94 / 49 / 31 %CV | Table 1 "IIV" columns |
| Residual error form | `ln(C_ij) = ln(C_pred,ij) + eps_ij` | Methods, same section |
| `propSd` / `propSd_pyrazol` | 34% / 20% | Table 1 "Residual variability" rows |
| Parent structure (2-cmt, 1st-order absorption + lag) | n/a | Methods, "Established pop PK model in healthy Beagle dogs" |
| Metabolite structure (3-cmt, 1st-order elimination) | n/a | Methods, same section |

## Reading Table 1: why `VM/F` shares `theta4` with `V/F`

Table 1 writes the metabolite central volume as `VM/F = theta4*(WGT/9.9)^1.0`
with `theta4 = 6.59 L` -- exactly the parent's `V/F` row. Two readings are
possible: the metabolite genuinely shares the parent's central-volume THETA, or
the table duplicated a row by mistake. Two independent checks select the first
reading.

**Theta accounting.** Table 1 names thetas 4, 5, 6, 7, 8, 9, 10, 11, 12 and 13.
The three unnamed low indices (1, 2, 3) are consumed by F, Tlag and ka, which are
the only parameters reported without a theta label. That accounts for thetas 1
through 13 with no gaps and nothing left over -- so no separate metabolite
central-volume theta exists in the model.

**Numerical check.** The metabolite's terminal half-life is strongly sensitive to
`VM/F`, and the value implied by `VM/F = 6.59 L` is reproduced below to three
significant figures against the paper's reported 13.8 days.

# Structural verification: terminal half-lives

The paper reports terminal half-lives of **1.4 days** for enflicoxib and **13.8
days** for the pyrazol metabolite (Results, "Enflicoxib and pyrazol metabolite
plasma levels"). Neither is a fitted parameter -- both are eigenvalues of the
disposition matrix -- so reproducing them is an independent test of the encoded
structure and the parameter roles.

The published values derive from the Beagle PK study, so the matrices below use
the Table 1 thetas as reported, i.e. at the 9.9 kg allometric reference weight.
(Half-life is weight-dependent in this model; see "Terminal half-life by washout
regression" below.)

```{r halflife-eigen}
wt_ref <- 9.9
p <- list(
  vc  = 6.59, vp = 27.70, q = 12.1, cl_met = 0.520, cl_nonmet = 0.193,
  vcm = 6.59, vpm = 29.3, vp2m = 13.4, qm = 13.2, q2m = 0.131, clm = 0.111
)

# Parent: 2-compartment disposition (absorption does not enter the terminal slope)
k10 <- (p$cl_met + p$cl_nonmet) / p$vc
k12 <- p$q / p$vc
k21 <- p$q / p$vp
A_parent <- matrix(c(-(k10 + k12), k21,
                     k12,         -k21), nrow = 2, byrow = TRUE)

# Metabolite: 3-compartment disposition
kme  <- p$clm / p$vcm
k12m <- p$qm  / p$vcm ; k21m <- p$qm  / p$vpm
k13m <- p$q2m / p$vcm ; k31m <- p$q2m / p$vp2m
A_met <- matrix(c(-(kme + k12m + k13m), k21m,  k31m,
                  k12m,                -k21m,  0,
                  k13m,                 0,    -k31m), nrow = 3, byrow = TRUE)

terminal_thalf_d <- function(A) log(2) / -max(Re(eigen(A)$values)) / 24

halflife_cmp <- tibble::tibble(
  Analyte     = c("Enflicoxib", "Pyrazol metabolite"),
  Model_days  = c(terminal_thalf_d(A_parent), terminal_thalf_d(A_met)),
  Published_d = c(1.4, 13.8)
) |>
  mutate(`Difference (%)` = 100 * (Model_days - Published_d) / Published_d)

halflife_cmp |>
  dplyr::rename("Terminal t1/2, model (d)" = Model_days,
                "Terminal t1/2, published (d)" = Published_d) |>
  knitr::kable(digits = 2,
               caption = "Terminal half-lives from the encoded disposition matrices vs Cendros 2025.")

# Gate: both half-lives within 5% of the published values.
stopifnot(nrow(halflife_cmp) == 2L,
          all(abs(halflife_cmp$`Difference (%)`) < 5))
```

Both reproduce within 3%. Because the metabolite half-life depends on `VM/F`,
the 13.81-vs-13.8-day agreement is what confirms the shared-`theta4` reading of
Table 1.

# Virtual cohort

Original observed data are not publicly available. The cohort below is a
virtual population of `r 83` dogs -- the size of the Cendros 2025 PK-evaluable
population -- whose body weights reproduce the published mean, SD and range.

The dosing regimen is the one Cendros 2025 used for its own external-evaluation
simulations (Results, "PK comparison between healthy Beagle dogs and dogs with
OA"): the **actual mean administered doses** of 10.4 mg/kg on day 0 followed by
5.2 mg/kg once weekly, rather than the nominal 8 / 4 mg/kg label doses.

```{r cohort}
set.seed(20250924)

n_dogs        <- 83L
wt_mean       <- 27.0    # Cendros 2025 Results, "Study population"
wt_sd         <- 15.0
wt_min        <- 4.9
wt_max        <- 64.9
dose_load     <- 10.4    # mg/kg, actual mean loading dose
dose_maint    <- 5.2     # mg/kg, actual mean maintenance dose
n_maint_doses <- 26L     # doses 2..27, weekly
tau_h         <- 7 * 24
study_end_h   <- 189 * 24

# Body weights: normal, truncated to the published range by resampling.
draw_wt <- function(n) {
  w <- stats::rnorm(n, wt_mean, wt_sd)
  while (any(bad <- w < wt_min | w > wt_max)) {
    w[bad] <- stats::rnorm(sum(bad), wt_mean, wt_sd)
  }
  w
}

subjects <- tibble::tibble(id = seq_len(n_dogs), WT = draw_wt(n_dogs),
                           regimen = "5.2 mg/kg weekly")

# Observation grid: 6 h through the study, refined to 1 h over the final
# steady-state dosing interval (days 182-189) where the NCA is computed.
obs_times <- sort(unique(c(
  seq(0, study_end_h, by = 6),
  seq(182 * 24, 189 * 24, by = 1)
)))

dose_times <- c(0, seq_len(n_maint_doses) * tau_h)
stopifnot(length(dose_times) == 27L, max(dose_times) == 182 * 24)

doses <- subjects |>
  tidyr::crossing(time = dose_times) |>
  mutate(amt  = ifelse(time == 0, dose_load, dose_maint) * WT,
         evid = 1L,
         cmt  = "depot")

# Observation rows carry the ODE state in `cmt` (never an observable name) and
# `dvid = 1L` to select an endpoint. The model declares two endpoints
# (Cc -> dvid 1, Cc_pyrazol -> dvid 2), and rxode2 requires observation rows to
# resolve to one of them. Both observables are returned as columns on every
# output row regardless of which dvid is selected, so one grid serves both.
obs <- subjects |>
  tidyr::crossing(time = obs_times) |>
  mutate(amt  = NA_real_,
         evid = 0L,
         cmt  = "central",
         dvid = 1L)

events <- bind_rows(doses, obs) |> arrange(id, time, desc(evid))

stopifnot(!anyDuplicated(events[, c("id", "time", "evid")]))

tibble::tibble(
  Statistic = c("N dogs", "Mean weight (kg)", "SD weight (kg)",
                "Min weight (kg)", "Max weight (kg)"),
  Simulated = c(n_dogs, mean(subjects$WT), sd(subjects$WT),
                min(subjects$WT), max(subjects$WT)),
  Published = c(83, 27.0, 15.0, 4.9, 64.9)
) |>
  knitr::kable(digits = 1, caption = "Virtual cohort vs Cendros 2025 demographics.")
```

# Simulation

```{r simulate}
mod <- readModelDb("Cendros_2025_enflicoxib")

sim <- rxode2::rxSolve(
  mod, events = events,
  keep       = c("WT", "regimen"),
  useLinCmt  = FALSE  # multi-analyte model: the ODE->linCmt auto-conversion breaks the mapping
) |>
  as.data.frame()
```

The typical-value profile at the cohort's mean weight (27 kg) is what Cendros
2025 simulated for its own external-evaluation figures, so it is reproduced
separately with the random effects zeroed.

```{r simulate-typical}
subj_typ <- tibble::tibble(id = 1L, WT = 27, regimen = "typical, 27 kg")

ev_typ <- bind_rows(
  subj_typ |> tidyr::crossing(time = dose_times) |>
    mutate(amt = ifelse(time == 0, dose_load, dose_maint) * WT, evid = 1L, cmt = "depot"),
  subj_typ |> tidyr::crossing(time = sort(unique(c(seq(0, study_end_h, by = 1))))) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L)
) |>
  arrange(time, desc(evid))

sim_typ <- rxode2::rxSolve(
  rxode2::zeroRe(mod), events = ev_typ, keep = c("WT", "regimen"), useLinCmt = FALSE
) |>
  as.data.frame()
```

# Replicate published results

## Figure 2 -- external evaluation of the Beagle model against the OA cohort

Cendros 2025 Figure 2 overlays the observed field-study concentrations on the
90% prediction interval simulated from the Beagle popPK model at 27 kg with the
actual mean doses. The equivalent prediction bands are below, with the paper's
observed means (Table 2) overlaid as points.

```{r figure-2}
bands <- sim |>
  select(id, time, Cc, Cc_pyrazol) |>
  tidyr::pivot_longer(c(Cc, Cc_pyrazol), names_to = "analyte", values_to = "conc") |>
  filter(!is.na(conc)) |>
  mutate(analyte = recode(analyte, Cc = "Enflicoxib", Cc_pyrazol = "Pyrazol metabolite")) |>
  group_by(analyte, time) |>
  summarise(Q05 = quantile(conc, 0.05), Q50 = quantile(conc, 0.50),
            Q95 = quantile(conc, 0.95), .groups = "drop")

# Cendros 2025 Table 2 observed means. The Table 2 cell for enflicoxib on day 44
# prints as "4,917", but the SD (462.5) and %CV (94.1) in the same row, and the
# Results/Discussion text ("491.7 and 262.5 ng/mL"), all identify the mean as
# 491.7 ng/mL. 462.5 / 491.7 = 94.1%. See "Assumptions and deviations".
observed <- tibble::tribble(
  ~analyte,             ~day, ~mean_obs,
  "Enflicoxib",           44,    491.7,
  "Enflicoxib",          189,    262.5,
  "Pyrazol metabolite",   44,   2105.6,
  "Pyrazol metabolite",  189,   2317.4
) |>
  mutate(time = day * 24)

ggplot(bands, aes(time / 24, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue") +
  geom_point(data = observed, aes(day, mean_obs), inherit.aes = FALSE,
             colour = "firebrick", size = 2.5) +
  facet_wrap(~analyte, scales = "free_y") +
  scale_y_log10() +
  labs(x = "Time (days)", y = "Plasma concentration (ng/mL)",
       title = "Figure 2 -- simulated 90% prediction interval vs observed means",
       caption = paste("Replicates Figure 2 of Cendros 2025. Band = 5th-95th percentile of the",
                       "83-dog virtual cohort; points = observed cohort means (Table 2)."))
```

## The paper's own simulated day-44 and day-189 concentrations

Cendros 2025 states its simulated mean enflicoxib concentrations were **765
ng/mL on day 44** and **118 ng/mL on day 189** (Results, "PK comparison between
healthy Beagle dogs and dogs with OA"). These are the sharpest available
numeric targets, because they are the paper's own model output rather than
observed data, so they test the encoding directly.

```{r paper-sim-check}
pick <- function(df, day, col) df[[col]][which.min(abs(df$time - day * 24))]

paper_sim <- tibble::tibble(
  Day             = c(44, 189),
  `Model (ng/mL)` = c(pick(sim_typ, 44, "Cc"), pick(sim_typ, 189, "Cc")),
  `Cendros 2025 simulated (ng/mL)` = c(765, 118),
  `Observed mean (ng/mL)`          = c(491.7, 262.5)
) |>
  mutate(`Difference vs paper simulation (%)` =
           100 * (`Model (ng/mL)` - `Cendros 2025 simulated (ng/mL)`) /
             `Cendros 2025 simulated (ng/mL)`)

knitr::kable(paper_sim, digits = 1,
             caption = "Enflicoxib typical-value predictions vs the paper's own simulated means.")

# Gate: reproduce the paper's own simulation to within 5%.
stopifnot(nrow(paper_sim) == 2L,
          all(abs(paper_sim$`Difference vs paper simulation (%)`) < 5))
```

The typical-value profile reproduces both of the paper's simulated
concentrations to within 1%. The paper's narrative around these numbers is also
recovered: the simulation runs above the observed mean on day 44 and below it on
day 189, which Cendros 2025 attributes to field-study variability and to the
27.7% of day-189 samples that fell below the limit of quantification.

## Figure 7 -- one year of weekly dosing shows no over-accumulation

Cendros 2025 Figure 7 simulates a full year of 4 mg/kg weekly administration to
demonstrate the absence of time-dependent PK. The paper's claim is specific:
enflicoxib reaches steady state by **week 4** and the pyrazol metabolite by
**weeks 10-12**, after which the profiles are constant.

```{r figure-7}
year_h <- 365 * 24
ev_year <- bind_rows(
  subj_typ |> tidyr::crossing(time = seq(0, year_h - 1, by = tau_h)) |>
    mutate(amt = 4 * WT, evid = 1L, cmt = "depot"),
  subj_typ |> tidyr::crossing(time = seq(0, year_h, by = 3)) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L)
) |>
  arrange(time, desc(evid))

sim_year <- rxode2::rxSolve(rxode2::zeroRe(mod), events = ev_year,
                            keep = "WT", useLinCmt = FALSE) |>
  as.data.frame()

sim_year |>
  select(time, Cc, Cc_pyrazol) |>
  tidyr::pivot_longer(-time, names_to = "analyte", values_to = "conc") |>
  mutate(analyte = recode(analyte, Cc = "Enflicoxib", Cc_pyrazol = "Pyrazol metabolite")) |>
  ggplot(aes(time / (7 * 24), conc, colour = analyte)) +
  geom_line() +
  labs(x = "Time (weeks)", y = "Plasma concentration (ng/mL)", colour = NULL,
       title = "Figure 7 -- one year of 4 mg/kg weekly dosing (typical 27 kg dog)",
       caption = "Replicates Figure 7 of Cendros 2025.") +
  theme(legend.position = "bottom")
```

```{r accumulation-gate}
# Peak concentration within each weekly dosing interval. The observation grid
# (every 3 h) divides the 168 h interval exactly, so every week is sampled at
# the same offsets relative to its own dose and the peaks are comparable.
weekly_peak <- sim_year |>
  mutate(week = floor(time / tau_h) + 1L) |>
  group_by(week) |>
  summarise(peak_parent = max(Cc), peak_met = max(Cc_pyrazol), .groups = "drop") |>
  filter(week <= 52)

stopifnot(nrow(weekly_peak) == 52L)

frac_of_plateau <- function(col, wk) {
  weekly_peak[[col]][weekly_peak$week == wk] / weekly_peak[[col]][weekly_peak$week == 52]
}

accum <- tibble::tibble(
  Week = c(4L, 12L, 26L),
  `Enflicoxib peak / week-52 peak` =
    vapply(c(4L, 12L, 26L), function(w) frac_of_plateau("peak_parent", w), numeric(1)),
  `Pyrazol peak / week-52 peak` =
    vapply(c(4L, 12L, 26L), function(w) frac_of_plateau("peak_met", w), numeric(1))
)

knitr::kable(accum, digits = 4, caption = paste(
  "Approach to plateau over one year of weekly dosing. Week 4 and weeks 10-12",
  "are the steady-state times Cendros 2025 reports; week 26 is the end of the",
  "field study."))

# Three separate claims from the paper, gated separately.
#
# 1. Enflicoxib is at steady state by week 4: its weekly peak from week 4
#    onward is indistinguishable (< 0.1%) from the week-52 peak.
stopifnot(abs(frac_of_plateau("peak_parent", 4L) - 1) < 0.001)

# 2. The pyrazol metabolite is at steady state by weeks 10-12: at week 12 it
#    is within 5% of its one-year plateau (the residual approach is the tail of
#    a 13.8-day half-life, and is upward, never an overshoot).
stopifnot(frac_of_plateau("peak_met", 12L) > 0.95,
          frac_of_plateau("peak_met", 12L) <= 1)

# 3. No over-accumulation: between the end of the 26-week field study and one
#    full year of continued weekly dosing, neither analyte's peak rises by
#    more than 0.5%.
stopifnot(abs(frac_of_plateau("peak_parent", 26L) - 1) < 0.005,
          abs(frac_of_plateau("peak_met", 26L) - 1) < 0.005)
```

Enflicoxib is flat from week 4 to within 0.1%, and the metabolite reaches 96% of
its one-year plateau by week 12 -- exactly the "4 weeks" and "10-12 weeks"
Cendros 2025 reports. Between the end of the 26-week study and a full year of
continued dosing neither analyte gains more than 0.5%, reproducing the paper's
central conclusion that weekly enflicoxib can be continued for as long as
therapeutically required.

The paper draws the same conclusion from a different observation -- that mean
pyrazol metabolite levels were "similar on days 44 and 189" -- which the model
also reproduces.

```{r time-independence}
met_d44  <- pick(sim_typ, 44,  "Cc_pyrazol")
met_d189 <- pick(sim_typ, 189, "Cc_pyrazol")

tibble::tibble(
  Day = c(44, 189),
  `Model pyrazol (ng/mL)`    = c(met_d44, met_d189),
  `Observed mean (ng/mL)`    = c(2105.6, 2317.4)
) |>
  knitr::kable(digits = 0, caption = paste(
    "Pyrazol metabolite on the two field-study sampling days: near-identical,",
    "consistent with steady state having been attained (Cendros 2025 Table 2)."))

# The model's day-44 and day-189 metabolite concentrations differ by under 2%,
# which is what "no time-dependent PK" means for this analyte.
stopifnot(abs(met_d189 / met_d44 - 1) < 0.02)
```

# PKNCA validation

NCA is computed on the final steady-state dosing interval (day 182 to day 189)
of the typical-value profile. The interval is shifted to a time origin of zero
so PKNCA sees a complete profile anchored by a dose at time 0 and a time-zero
concentration record.

```{r pknca-setup}
ss_start_h <- 182 * 24

ss_typ <- sim_typ |>
  filter(time >= ss_start_h, time <= ss_start_h + tau_h) |>
  mutate(time = time - ss_start_h, id = 1L, regimen = "5.2 mg/kg weekly")

stopifnot(nrow(ss_typ) > 0, min(ss_typ$time) == 0, max(ss_typ$time) == tau_h)

ss_dose <- tibble::tibble(id = 1L, time = 0, amt = dose_maint * 27,
                          regimen = "5.2 mg/kg weekly")

intervals_ss <- data.frame(
  start = 0, end = tau_h,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE
)

run_nca <- function(conc_col) {
  df <- ss_typ |>
    select(id, time, regimen, conc = dplyr::all_of(conc_col)) |>
    filter(!is.na(conc))
  cobj <- PKNCA::PKNCAconc(df, conc ~ time | regimen + id)
  dobj <- PKNCA::PKNCAdose(ss_dose, amt ~ time | regimen + id)
  PKNCA::pk.nca(PKNCA::PKNCAdata(cobj, dobj, intervals = intervals_ss))
}

nca_parent <- run_nca("Cc")
nca_met    <- run_nca("Cc_pyrazol")
```

## Closed-form identity check

At steady state the area under a dosing interval must equal the dose delivered
into the compartment divided by the clearance out of it. This identity gates the
ODE system, the dose encoding, the interval window **and** PKNCA's settings
simultaneously -- if any one of them is wrong, it fails.

```{r closed-form}
wt   <- 27
allo <- (wt / wt_ref)^0.75

cl_parent_tot <- (p$cl_met + p$cl_nonmet) * allo    # L/h
cl_met_ind    <- p$cl_met * allo                    # L/h, formation arm
cl_met_elim   <- p$clm * allo                       # L/h, metabolite elimination
dose_mg       <- dose_maint * wt

# AUC in ng*h/mL: (mg / (L/h)) = mg*h/L = ug*h/mL, x 1000 -> ng*h/mL
auc_parent_cf <- dose_mg / cl_parent_tot * 1000
fm            <- p$cl_met / (p$cl_met + p$cl_nonmet)
auc_met_cf    <- (fm * dose_mg) / cl_met_elim * 1000

auc_nca <- function(res) {
  s <- as.data.frame(res$result)
  s$PPORRES[s$PPTESTCD == "auclast"]
}

cf <- tibble::tibble(
  Analyte = c("Enflicoxib", "Pyrazol metabolite"),
  `PKNCA AUCtau (ng*h/mL)`     = c(auc_nca(nca_parent), auc_nca(nca_met)),
  `Closed form (ng*h/mL)`      = c(auc_parent_cf, auc_met_cf)
) |>
  mutate(`Difference (%)` = 100 * (`PKNCA AUCtau (ng*h/mL)` - `Closed form (ng*h/mL)`) /
           `Closed form (ng*h/mL)`)

knitr::kable(cf, digits = c(0, 0, 0, 2),
             caption = "Steady-state AUCtau from PKNCA vs the closed-form dose/clearance identity.")

stopifnot(nrow(cf) == 2L, all(abs(cf$`Difference (%)`) < 1))
```

Both analytes match their closed-form value to better than 1%. For the
metabolite this also confirms the fraction metabolised implied by the two parent
elimination arms, `fm = CL2/(CL1+CL2) = `r round(fm, 3)``.

## Comparison against published exposures

Cendros 2025 reports mean individual Bayesian exposure estimates at steady state
(Results, "Bayesian approach ...", drawn from Supplementary Figures S3 and S4):
enflicoxib Cmax ~1,173 ng/mL, Cmin ~185 ng/mL and AUCtau ~3,810 ng/mL*day; the
pyrazol metabolite ~2,490 ng/mL, ~2,243 ng/mL and ~16,742 ng/mL*day over weeks
12-26.

```{r nca-compare}
to_day <- function(res, analyte) {
  s <- as.data.frame(res$result)
  tibble::tibble(
    analyte    = analyte,
    cmax       = s$PPORRES[s$PPTESTCD == "cmax"],
    cmin       = s$PPORRES[s$PPTESTCD == "cmin"],
    auclast    = s$PPORRES[s$PPTESTCD == "auclast"] / 24   # ng*h/mL -> ng*d/mL
  )
}

simulated <- bind_rows(to_day(nca_parent, "Enflicoxib"),
                       to_day(nca_met, "Pyrazol metabolite"))

published <- tibble::tribble(
  ~analyte,             ~cmax,  ~cmin,  ~auclast,
  "Enflicoxib",          1173,    185,      3810,
  "Pyrazol metabolite",  2490,   2243,     16742
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = simulated,
  reference     = published,
  by            = "analyte",
  units         = c(cmax = "ng/mL", cmin = "ng/mL", auclast = "ng*d/mL"),
  tolerance_pct = 20
)

knitr::kable(cmp, caption = paste(
  "Typical-value steady-state NCA vs the mean individual Bayesian estimates",
  "reported by Cendros 2025. * differs from the reference by >20%."))
```

`AUCtau` agrees to 1.5% for the parent and 8% for the metabolite, and `Cmax` to
within 13% for both. The one starred row is the parent's **Cmin**, which the
typical-value profile places below the reported cohort mean.

This is expected rather than a defect. Cmin is compared against a *mean of
individual Bayesian estimates*, and the parent carries very large log-normal
IIV (94% CV on V/F, 101% on ka, 49% on CL2/F). Under a log-normal random effect
the cohort mean sits above the typical value by roughly `exp(omega^2/2)`, and
the effect is largest exactly where the paper observed it: Cendros 2025 notes
that Cmin showed "an increase of interindividual variability ... with a higher
right-skewed distribution for Cmin". The stochastic cohort makes the point:

```{r cmin-skew}
ss_cohort <- sim |>
  filter(time >= ss_start_h, time <= ss_start_h + tau_h) |>
  group_by(id) |>
  summarise(cmin_parent = min(Cc), cmax_parent = max(Cc), .groups = "drop")

tibble::tibble(
  Statistic = c("Typical-value Cmin", "Cohort median Cmin", "Cohort mean Cmin",
                "Published mean Cmin"),
  `ng/mL`   = c(simulated$cmin[simulated$analyte == "Enflicoxib"],
                median(ss_cohort$cmin_parent),
                mean(ss_cohort$cmin_parent),
                185)
) |>
  knitr::kable(digits = 0,
               caption = "Enflicoxib steady-state Cmin: typical value vs a right-skewed cohort.")

# The cohort mean must exceed the cohort median (right skew) and must sit
# above the typical value -- the mechanism that explains the Cmin gap.
stopifnot(mean(ss_cohort$cmin_parent) > median(ss_cohort$cmin_parent),
          mean(ss_cohort$cmin_parent) >
            simulated$cmin[simulated$analyte == "Enflicoxib"])
```

## Terminal half-life by washout regression

A single dose followed by a long washout recovers the terminal half-lives by
log-linear regression, independently of the eigenvalue calculation above.

The published values of 1.4 and 13.8 days come from the Beagle PK study (Homedes
2021, Cendros 2025 reference 4), so the washout is simulated at the **9.9 kg
allometric reference weight**, not at the OA cohort's 27 kg mean. That
distinction matters: clearances scale as `WT^0.75` while volumes scale as
`WT^1.0`, so every rate constant scales as `WT^-0.25` and half-life scales as
`WT^+0.25`. A 27 kg dog therefore has a genuinely longer half-life than a 9.9 kg
Beagle, by a factor the model must reproduce exactly.

```{r halflife-washout}
simulate_washout <- function(wt) {
  s <- tibble::tibble(id = 1L, WT = wt)
  ev <- bind_rows(
    s |> mutate(time = 0, amt = dose_maint * wt, evid = 1L, cmt = "depot"),
    s |> tidyr::crossing(time = seq(0, 150 * 24, by = 3)) |>
      mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L)
  ) |>
    arrange(time, desc(evid))
  rxode2::rxSolve(rxode2::zeroRe(mod), events = ev, keep = "WT",
                  useLinCmt = FALSE) |>
    as.data.frame()
}

# Log-linear slope over a window well past the distribution phase. Rows where
# the concentration has decayed to zero are dropped so log() stays finite.
slope_thalf_d <- function(df, col, from_d, to_d) {
  w <- df[df$time >= from_d * 24 & df$time <= to_d * 24, ]
  w <- w[w[[col]] > 0, ]
  stopifnot(nrow(w) >= 10)
  log(2) / -stats::coef(stats::lm(log(w[[col]]) ~ w$time))[[2]] / 24
}

wash_ref <- simulate_washout(wt_ref)   # 9.9 kg, the allometric reference
wash_27  <- simulate_washout(27)       # the OA cohort's mean weight

hl_parent_ref <- slope_thalf_d(wash_ref, "Cc",         5, 12)
hl_met_ref    <- slope_thalf_d(wash_ref, "Cc_pyrazol", 50, 110)

hl_nca <- tibble::tibble(
  Analyte = c("Enflicoxib", "Pyrazol metabolite"),
  `Regression t1/2 at 9.9 kg (d)` = c(hl_parent_ref, hl_met_ref),
  `Published t1/2 (d)`            = c(1.4, 13.8)
) |>
  mutate(`Difference (%)` = 100 * (`Regression t1/2 at 9.9 kg (d)` - `Published t1/2 (d)`) /
           `Published t1/2 (d)`)

knitr::kable(hl_nca, digits = 3, caption = paste(
  "Terminal half-life by log-linear regression on a single-dose washout at the",
  "9.9 kg allometric reference weight."))

stopifnot(nrow(hl_nca) == 2L, all(abs(hl_nca$`Difference (%)`) < 5))
```

The same regression applied at 27 kg tests the allometric exponents themselves.
Because half-life scales as `WT^(1 - 0.75) = WT^0.25`, the ratio of the two
weights' half-lives must equal `(27/9.9)^0.25` for **both** analytes -- a
prediction with no free parameters that would break if the 0.75 and 1.0
exponents were swapped, shared, or applied to the wrong parameter class.

```{r allometry-gate}
hl_parent_27 <- slope_thalf_d(wash_27, "Cc",         8, 20)
hl_met_27    <- slope_thalf_d(wash_27, "Cc_pyrazol", 60, 120)

expected_ratio <- (27 / wt_ref)^(1 - 0.75)

allo_cmp <- tibble::tibble(
  Analyte = c("Enflicoxib", "Pyrazol metabolite"),
  `t1/2 at 9.9 kg (d)` = c(hl_parent_ref, hl_met_ref),
  `t1/2 at 27 kg (d)`  = c(hl_parent_27, hl_met_27)
) |>
  mutate(`Observed ratio` = `t1/2 at 27 kg (d)` / `t1/2 at 9.9 kg (d)`,
         `Predicted (27/9.9)^0.25` = expected_ratio)

knitr::kable(allo_cmp, digits = 4, caption = paste(
  "Half-life scaling with body weight. The ratio is fixed by the difference",
  "between the clearance exponent (0.75) and the volume exponent (1.0)."))

stopifnot(all(abs(allo_cmp$`Observed ratio` - expected_ratio) < 1e-3))
```

Both analytes scale by `r sprintf("%.4f", expected_ratio)`, matching the
prediction to better than one part in a thousand.

# Assumptions and deviations

- **Parameters are the Cendros 2022 estimates.** Every `ini()` value comes from
  Cendros 2025 Table 1, which reproduces the model estimated by Cendros 2022 in
  healthy Beagle dogs. Cendros 2025 re-estimated nothing; it performed an
  external evaluation (VPC / pcVPC / NPDE) plus a MAP-Bayesian `POSTHOC` fit.
  The upstream Cendros 2022 paper was not consulted directly -- it was not
  required, because Table 1 and its two footnotes give every parameter value,
  every parameter's structural role, the allometric equation, the IIV
  distribution and the residual-error form.

- **`VM/F` shares `theta4` with the parent `V/F`.** Table 1 assigns the same
  theta and the same value (6.59 L) to both. This is treated as the model's real
  structure rather than a table slip, on two grounds set out in "Reading Table 1"
  above: thetas 1-13 are fully accounted for with no spare index for a separate
  metabolite volume, and the metabolite terminal half-life implied by 6.59 L
  reproduces the published 13.8 days to three significant figures.

- **`theta9` (VMP/F) prints with a "-" in Table 1's Units column.** The sibling
  volume rows (`theta4`, `theta13`) print "L" and VMP/F is defined in the Table 1
  footnote as "apparent volume of distribution in the shallow peripheral
  compartment", so the units are taken as litres. This is a formatting omission
  in the table, not an ambiguity about the quantity.

- **Table 2's day-44 enflicoxib mean is misprinted.** The cell reads "4,917"
  ng/mL, but the same row's SD (462.5) and %CV (94.1) are consistent only with a
  mean of 491.7, and the Results and Discussion text both give the value as
  491.7 / 492 ng/mL. The vignette uses 491.7.

- **Residual error encoded as proportional.** The paper's residual model is
  additive on the natural-log scale (`ln C = ln C_pred + eps`), which is
  proportional error on the linear scale in nlmixr2, and its magnitude is
  reported as a %CV. The %CV is carried directly as `propSd`; for these
  magnitudes the exact log-scale SD differs by under 3%
  (`sqrt(log(1 + 0.34^2)) = 0.331` vs 0.34).

- **Omega matrix assumed diagonal.** Table 1 reports a single %CV per parameter
  and no correlations or covariances between random effects, so no off-diagonal
  terms are encoded.

- **IIV on a bioavailability fixed at 1.** Table 1 reports F as "1 FIX" with 39%
  IIV. This is encoded faithfully as `lfdepot <- fixed(log(1))` with
  `etalfdepot`, which means individual F values exceed 1 for positive etas. That
  is a property of the published model (F is a *relative* bioavailability
  absorbed into every apparent parameter), not of this encoding.

- **Metabolite input is a mass flux.** The metabolite ODE is fed by
  `cl_met / vc * central` with no molar-mass correction. Every metabolite
  parameter in Table 1 is an apparent "/F" quantity, so the molar-mass ratio and
  the metabolite's own availability are absorbed into `VM/F`, `CLM/F` and the
  rest; no separate conversion factor is published or needed.

- **Body-weight distribution assumed normal.** Cendros 2025 reports the cohort
  weight as mean 27.0 +/- 15 kg over a 4.9-64.9 kg range but does not publish the
  distribution. A normal distribution truncated to that range by resampling is
  used. Age and sex were screened by the paper and not retained, so they are
  recorded in `covariatesDataExcluded` and are not simulated.

- **Doses are the actual mean administered doses.** Simulations use 10.4 mg/kg
  loading and 5.2 mg/kg maintenance, which is what Cendros 2025 used for its own
  external-evaluation figures, rather than the 8 / 4 mg/kg label doses. The
  one-year Figure 7 replication uses the paper's 4 mg/kg for that figure.

- **Observed data are not reproduced.** Individual concentrations from the field
  study are not public. Comparisons are made against the paper's published
  summary statistics (Table 2 means) and against its own simulated and Bayesian
  exposure estimates.
