---
title: "Sunitinib thrombocytopenia (Centanni 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sunitinib thrombocytopenia (Centanni 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(ggplot2)
```

# Model and source

* Citation: Centanni M, Nijhuis J, Karlsson MO, Friberg LE. Comparative
  Analysis of Traditional and Pharmacometric-Based Pharmacoeconomic
  Modeling in the Cost-Utility Evaluation of Sunitinib Therapy.
  *PharmacoEconomics* 2025;43(1):31-43.
* Article: [doi:10.1007/s40273-024-01438-z](https://doi.org/10.1007/s40273-024-01438-z)
* Parameters: Online Resource 1 (Springer electronic supplementary
  material), section "Absolute Thrombocyte Count Model", **Table S2**.
* Parent framework (not re-estimated by Centanni 2025):
  `Hansson_2013_sunitinib_myelosuppression` (ANC),
  `Hansson_2013_sunitinib_dbp`, `Hansson_2013_sunitinib_hfs`,
  `Hansson_2013_sunitinib_os`,
  [doi:10.1038/psp.2013.62](https://doi.org/10.1038/psp.2013.62).
* Friberg myelosuppression backbone: Friberg LE et al. *J Clin Oncol*
  2002;20(24):4713-4721,
  [doi:10.1200/JCO.2002.02.140](https://doi.org/10.1200/JCO.2002.02.140).

Centanni 2025 is a health-economics methods paper: it compares a
pharmacometric-based pharmacoeconomic framework against four traditional
pharmacoeconomic frameworks for a cost-utility analysis of sunitinib in
gastrointestinal stromal tumour (GIST). Most of its pharmacometric
content is the previously published Hansson 2013 framework, which is
already packaged in nlmixr2lib under the model names listed above.

**One component of the framework is original to this paper**, and it is
what this vignette packages: a semi-physiological Friberg-Karlsson
myelosuppression model for the **absolute thrombocyte (platelet) count**.
The main text is explicit that this is the sole newly estimated
component (Methods 2.2: "The pharmacometric-based models were not
re-estimated, except for the thrombocytopenia model, which was newly
developed"). No sunitinib platelet model was previously in nlmixr2lib.

The paper's four traditional pharmacoeconomic model structures
(time-to-event exponential and Weibull, discrete-time Markov,
continuous-time Markov) and its five logistic-regression toxicity models
are given as structural equations only (Online Resource 1, Equations
1-12) with **no parameter estimates reported anywhere**, and they were
fitted to simulated rather than observed data. They are therefore not
extractable and are not packaged.

# Population

Centanni 2025 simulated a two-arm trial of N = 1,000 virtual GIST
patients comparing sunitinib 37.5 mg orally once daily (continuous
schedule, with protocol dose reductions to 25, 12.5 or 0 mg on
unacceptable adverse events) against no treatment. Virtual-patient
covariates were generated from the distributions described in the main
text section 2.1; the paper does not report an observed
baseline-demographics table.

Platelet count was evaluated every 6 weeks in the simulation, in line
with established clinical protocols, and **unacceptable thrombocytopenia
was defined as a platelet count below 50 x 10^9/L**, aggregated per
6-week cycle (Methods 2.1).

Note on provenance: the N = 1,000 above is the size of the *simulated*
trial the pharmacoeconomic comparison was run on, not an observed
estimation cohort. The source does not state which patient dataset the
thrombocyte model itself was estimated on; by construction of the
framework it is the pooled four-study imatinib-resistant GIST sunitinib
dataset (n = 303) underlying Hansson 2013. The model's `population`
metadata records this explicitly.

```{r population}
str(rxode2::rxode(readModelDb("Centanni_2025_sunitinib_thrombocytopenia"))$population)
```

# Errata: a suspected factor-of-1000 scale error in Table S2

**Three of the five structural parameters in this model are encoded at
1/1000 of the value printed in the source table.** This is the single
most important thing to know about this extraction, so it is stated
before the source trace rather than buried in a closing section.

Online Resource 1 Table S2 prints, verbatim:

| Parameter | Estimate | RSE, % | IIV CV, % | RSE, % |
|-----------|----------|--------|-----------|--------|
| Baseline thrombocyte count (x 10^9/liter) | 323 | 2.3 | 33.3 | 5 |
| MTT (hours) | 175 | 4.2 | 17.7 | 9.4 |
| Emax | 497 | 33.2 | - | - |
| EC50 (ng * hour / L) | 10300 | 34.6 | 66.9 | 11.3 |
| gamma | 92.3 | 8.6 | | |
| Residual error (x 10^9/liter) | 0.268 | 6% | | |

`Emax = 497` and `gamma = 92.3` cannot be reconciled with the Friberg
structure that Table S2's own accompanying text points to, and the
`EC50` unit `ng * hour / L` is inconsistent with the exposure scale the
framework uses. The three corrections are the **same factor of 1000 in
the same direction within one table**, which is consistent with a single
systematic scale/unit error in the typesetting of Table S2 rather than
with three independent estimation results. For comparison, the
corrected `Emax` of 0.497 sits essentially on top of the sibling ANC
model's `Emax` of 0.520 (Hansson 2013 Table 2), and the corrected
`gamma` of 0.0923 is the same order as that model's `gamma` of 0.362.

This is described as a suspected erratum in the source, **not** as an
error by the authors: the printed values are internally identifiable as
a scale slip, and every other value in the table is used exactly as
printed.

## Falsification of the as-printed values

The claim above is testable, and the test is run here rather than
asserted. Each variant below substitutes the as-printed value(s) back
into the packaged model and solves a typical-value patient on the
paper's own regimen (37.5 mg/day, 26 weeks).

```{r falsify, message = FALSE, warning = FALSE}
mod <- readModelDb("Centanni_2025_sunitinib_thrombocytopenia")

fals_times  <- seq(0, 26 * 7 * 24, by = 24)
fals_events <- data.frame(
  id = 1L, time = fals_times, evid = 0L, amt = 0, cmt = "circ",
  DOSE = 37.5, CLI = 32.819
)

solve_variant <- function(...) {
  changes <- list(...)
  m <- if (length(changes)) do.call(rxode2::ini, c(list(mod), changes)) else mod
  out <- try(
    as.data.frame(rxode2::rxSolve(rxode2::zeroRe(m), fals_events,
                                  atol = 1e-8, rtol = 1e-8)),
    silent = TRUE
  )
  if (inherits(out, "try-error")) return(NA_real_)
  tail(out$circ, 1)
}

falsification <- tibble::tibble(
  Reading = c(
    "Encoded here (Emax 0.497, EC50 10.3 mg*h/L, gamma 0.0923)",
    "gamma as printed (92.3)",
    "Emax as printed (497)",
    "EC50 unit as printed (10300 ng*h/L = 0.0103 mg*h/L)",
    "All three as printed"
  ),
  platelets_wk26 = c(
    solve_variant(),
    solve_variant(lgamma = log(92.3)),
    solve_variant(lemax  = log(497)),
    solve_variant(lec50  = log(0.0103)),
    solve_variant(lemax = log(497), lec50 = log(0.0103), lgamma = log(92.3))
  )
) |>
  # formatC, not kable(digits=), so that 2e-25 does not print as 0.000
  mutate(`Platelets at week 26 (10^9/L)` =
           formatC(platelets_wk26, format = "g", digits = 4),
         Verdict = c(
           "Stable, physiologically plausible",
           "Diverges (unbounded growth)",
           "Complete marrow ablation",
           "Near-complete ablation",
           "Unsolvable (NaN)"
         )) |>
  select(-platelets_wk26)

knitr::kable(falsification)
```

Only the encoded reading is solvable and plausible. `gamma = 92.3` is a
very strong feedback gain that drives the seven-stage chain outside its
stable region; `Emax = 497` and the as-printed `EC50` unit both drive
the proliferation term to (or past) total suppression, which is
irreconcilable with the roughly 3% per-cycle thrombocytopenia incidence
the paper itself reports.

Encoded per the operator ruling on task sidecar `oare_PMC11724784`
request-001 (2026-08-14).

# Source trace

The source gives this model's structure **only by reference**. Online
Resource 1 states, in full:

> "A semi-physiological myelosuppression model was developed to describe
> changes in absolute thrombocyte count following sunitinib treatment
> (Table S2). The structural model is similar to the published ANC
> model, with the exception that five transit compartment were used
> instead of three. Additionally, sunitinib area under the curve (AUC)
> was the best predictor of thrombocyte changes. An Emax function
> described the relationship between AUC and thrombocytes."

No equation is printed. The structure below is therefore the Hansson
2013 ANC model's published structure (packaged as
`Hansson_2013_sunitinib_myelosuppression`) with the two stated
modifications applied: five transit compartments instead of three, and
sunitinib AUC as the Emax driver in place of `sVEGFR-3 REL`.

| Equation / parameter | Source location |
|----------------------|------------------|
| Friberg chain, 7 states (prol + 5 transit + circ) | Online Resource 1: "similar to the published ANC model ... five transit compartment were used instead of three" |
| `ktr <- 6 / mtt` (Friberg n = 5 plus prol form) | Friberg 2002 convention with n_transit = 5, per the sentence above |
| `edrug <- emax * auc / (ec50 + auc)` | Online Resource 1: "An Emax function described the relationship between AUC and thrombocytes" |
| `auc <- DOSE / CLI` | Hansson 2013 framework convention for the per-cycle exposure summary; inherited unchanged (the PK was not re-estimated) |
| `feed <- (circ0 / circ)^gamma` | Inherited from the ANC model; Table S2 footnote defines gamma as the "feedback factor" |
| `lcirc0 = log(323)` | Table S2 "Baseline thrombocyte count (x 10^9/liter)" = 323 (RSE 2.3%) |
| `lmtt = log(175)` | Table S2 "MTT (hours)" = 175 (RSE 4.2%) |
| `lemax = log(0.497)` | Table S2 "Emax" printed **497** (RSE 33.2%); suspected 1000x scale erratum |
| `lec50 = log(10.3)` | Table S2 "EC50 (ng * hour / L)" printed **10300** (RSE 34.6%); read as 10300 ng*h/mL = 10.3 mg*h/L |
| `lgamma = log(0.0923)` | Table S2 "gamma" printed **92.3** (RSE 8.6%); suspected 1000x scale erratum |
| `etalcirc0 ~ 0.105161` | Table S2 Circ0 IIV CV 33.3%; omega^2 = log(0.333^2 + 1) |
| `etalmtt ~ 0.030848` | Table S2 MTT IIV CV 17.7%; omega^2 = log(0.177^2 + 1) |
| `etalec50 ~ 0.369880` | Table S2 EC50 IIV CV 66.9%; omega^2 = log(0.669^2 + 1) |
| `addSd = 0.268` | Table S2 "Residual error (x 10^9/liter)" = 0.268 (RSE 6%) |

The IIV back-transformation follows the Table S2 footnote, which defines
the reported CV as `sqrt(exp(omega^2) - 1)`, so
`omega^2 = log(CV^2 + 1)`. `Emax` and `gamma` are reported with "-" in
the IIV column and carry no eta. Table S2 reports no correlations, so
the omega matrix is diagonal (unlike the sibling ANC model, where
Hansson 2013 reported a 90% ANC0-Emax correlation).

## Units

| Symbol | Units | Note |
|--------|-------|------|
| `DOSE` | mg | daily sunitinib dose, data column |
| `CLI` | L/h | individual upstream popPK clearance, data column |
| `auc` | mg*h/L | `DOSE / CLI` = mg / (L/h) |
| `ec50` | mg*h/L | same scale as `auc`, so `edrug` is unitless |
| `emax` | unitless | fraction of proliferation inhibited |
| `gamma` | unitless | exponent on the unitless ratio `circ0 / circ` |
| `mtt` | h | `ktr = 6 / mtt` therefore has units 1/h |
| `prol`, `transit1..5`, `circ` | 10^9 cells/L | `d/dt(state)` = (1/h) * (10^9 cells/L), consistent |

Every ODE term is `ktr * <state>` (times unitless factors), giving
`10^9 cells/L/h` on both sides. The only non-obvious conversion is the
`EC50` one discussed in the Errata section.

# Required covariates

This model has **no PK ODE**. Sunitinib exposure enters as the daily AUC
summary `auc = DOSE / CLI`, so two data columns are required:

* `DOSE` (mg) -- daily sunitinib dose; set to 0 during an interruption
  or for an untreated subject, which sets `auc` to 0 and removes the
  drug effect.
* `CLI` (L/h) -- per-subject sunitinib clearance from the upstream
  popPK model (Houk 2009, not packaged in nlmixr2lib). Use
  32.819 L/h for typical-value simulations, matching the convention of
  every sibling model in the Hansson 2013 framework.

# Validation

Because the output is a circulating cell count with no dosing or
absorption-distribution-elimination profile, NCA is not the right
validation target. The checks below follow the steady-state /
perturbation-recovery / mass-balance pattern used for turnover models,
plus a replication of the paper's own reported thrombocytopenia
incidence.

```{r validation-setup}
mod  <- readModelDb("Centanni_2025_sunitinib_thrombocytopenia")
modT <- rxode2::zeroRe(mod)

cycle_h   <- 6 * 7 * 24          # the paper's 6-week evaluation cycle
horizon_h <- 104 * 7 * 24        # the paper's 104-week horizon
tv_cl     <- 32.819              # framework typical clearance (L/h)

make_events <- function(times, dose, cli = tv_cl, id = 1L) {
  data.frame(id = id, time = times, evid = 0L, amt = 0, cmt = "circ",
             DOSE = dose, CLI = cli)
}
```

## 1. Steady-state hold with no drug

With `DOSE = 0` the chain must sit at the reported baseline
`Circ0 = 323 x 10^9/L` indefinitely. This is an exact check: any sign
error, missing term, or mis-set initial condition breaks it.

```{r ss-hold}
t_daily <- seq(0, horizon_h, by = 24)
sim_off <- rxode2::rxSolve(modT, make_events(t_daily, dose = 0)) |>
  as.data.frame()

range(sim_off$circ)
stopifnot(all(abs(sim_off$circ - 323) < 1e-6))
```

Held at 323 to within 1e-6 over the full 104-week horizon.

## 2. Typical-value steady state against the closed form

At steady state `d/dt(prol) = 0` requires
`(1 - edrug) * (Circ0 / circ)^gamma = 1`, which rearranges to a closed
form independent of `MTT` and of the number of transit compartments:

`circ_ss / Circ0 = (1 - edrug)^(1 / gamma)`

This is the strongest single gate in the vignette: it pins `Circ0`,
`Emax`, `EC50`, `gamma` and the feedback structure simultaneously
against a value the solver never sees.

```{r ss-closed-form}
emax_tv  <- 0.497
ec50_tv  <- 10.3
gamma_tv <- 0.0923
circ0_tv <- 323

auc_tv   <- 37.5 / tv_cl
edrug_tv <- emax_tv * auc_tv / (ec50_tv + auc_tv)
ss_closed_form <- circ0_tv * (1 - edrug_tv)^(1 / gamma_tv)

sim_on <- rxode2::rxSolve(modT, make_events(t_daily, dose = 37.5)) |>
  as.data.frame()
ss_ode <- tail(sim_on$circ, 1)

c(daily_auc_mg_h_L = auc_tv, edrug = edrug_tv,
  closed_form = ss_closed_form, ode = ss_ode)

stopifnot(abs(ss_ode - ss_closed_form) / ss_closed_form < 1e-5)
```

The ODE steady state agrees with the closed form to better than 1 part
in 1e5. A typical patient on 37.5 mg/day settles at about
`r round(ss_closed_form)` x 10^9/L, a `r round(100 * (1 - ss_closed_form / circ0_tv))`%
reduction from baseline -- a clinically sensible sunitinib
thrombocytopenia signal, and the reason a minority of patients cross the
50 x 10^9/L threshold.

## 3. Perturbation recovery on dose interruption

The paper's protocol reduces the dose to 0 mg on a grade-3 event and
reinstates it after recovery. That makes dose interruption the natural
perturbation test: platelets should fall under drug, then return to
`Circ0` once the drug is withdrawn.

```{r recovery}
t_rec    <- seq(0, 78 * 7 * 24, by = 24)
dose_rec <- ifelse(t_rec <= 26 * 7 * 24, 37.5, 0)   # 26 weeks on, then off

sim_rec <- rxode2::rxSolve(modT, make_events(t_rec, dose = dose_rec)) |>
  as.data.frame()

c(nadir           = min(sim_rec$circ),
  at_interruption = sim_rec$circ[sim_rec$time == 26 * 7 * 24],
  final           = tail(sim_rec$circ, 1))

stopifnot(min(sim_rec$circ) < 323)                      # drug depresses platelets
stopifnot(abs(tail(sim_rec$circ, 1) - 323) < 1e-3)      # full recovery to baseline
```

```{r recovery-fig, fig.width = 7, fig.height = 3.5}
ggplot(sim_rec, aes(time / (7 * 24), circ)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 323, linetype = "dashed", colour = "grey50") +
  geom_hline(yintercept = 50, linetype = "dotted", colour = "firebrick") +
  geom_vline(xintercept = 26, linetype = "dotted", colour = "grey30") +
  annotate("text", x = 27, y = 300, hjust = 0, size = 3,
           label = "dose interrupted") +
  labs(x = "Time (weeks)", y = "Platelets (10^9/L)",
       title = "Typical-value platelet trajectory: 26 weeks on 37.5 mg/day, then withdrawal",
       subtitle = "Dashed = baseline 323; dotted red = 50 x 10^9/L thrombocytopenia threshold") +
  theme_minimal()
```

## 4. Mass balance at steady state

At steady state under drug every transfer in the chain must balance. The
checks below are read off the **solved** system (the section 2 solve), so
they test the ODE the solver actually integrated rather than restating an
algebraic identity.

```{r mass-balance}
mtt_tv <- 175
ktr_tv <- 6 / mtt_tv

ss     <- tail(sim_on, 1)
states <- unlist(ss[c("prol", "transit1", "transit2", "transit3",
                      "transit4", "transit5", "circ")])
states

# 1. A uniform-ktr chain equilibrates every stage to the SAME level, because
#    d/dt(transit_j) = ktr * (transit_{j-1} - transit_j) = 0. A dropped or
#    mis-wired transfer term shows up here immediately.
stopifnot(max(states) / min(states) - 1 < 1e-9)

# 2. Flux balance across the whole chain: production into `prol` must equal
#    elimination out of `circ`. Both sides are computed from the solved state
#    values, so this is not an arithmetic tautology.
feed_ss  <- (circ0_tv / ss$circ)^gamma_tv
prol_in  <- ktr_tv * ss$prol * (1 - edrug_tv) * feed_ss
circ_out <- ktr_tv * ss$circ
c(prol_in = prol_in, circ_out = circ_out)
stopifnot(abs(prol_in / circ_out - 1) < 1e-9)

# 3. The feedback identity that a non-zero steady state implies:
#    (1 - edrug) * (Circ0/circ)^gamma = 1.
stopifnot(abs((1 - edrug_tv) * feed_ss - 1) < 1e-9)
```

All seven states equilibrate to `r sprintf("%.1f", states[["circ"]])` x
10^9/L, the chain influx matches the circulating efflux to machine
precision, and the feedback identity closes.

## 5. Reported thrombocytopenia incidence (replicates Fig. 3)

Section 3.1 of the paper reports, for the pharmacometric framework at
sunitinib 37.5 mg/day with thrombocytopenia defined as platelets
< 50 x 10^9/L aggregated per 6-week cycle:

* highest incidence in any single treatment cycle: **3.0%**
* average presence over the treatment period: **0.75%**

and Fig. 3 shows the per-cycle time course, with the pharmacometric
framework's toxicities *declining over time* -- the paper's central
qualitative contrast against the traditional frameworks, where time has
no influence.

These two incidence figures come from the paper's 10,000-patient
cost-utility cohort (Fig. 3 caption; the N = 1,000 two-arm trial in the
Population section above is the earlier simulation used to estimate the
traditional pharmacoeconomic models). Fig. 3 expresses incidence as a
percentage of the patients *alive* in each cycle.

`Circ0` enters the system as a pure multiplicative scale, so
`circ_i(t) = Circ0_i * h(t; MTT_i, EC50_i)`. That lets the `Circ0`
lognormal be integrated **analytically** and only the `MTT` and `EC50`
etas be handled by quadrature, giving an essentially noise-free
incidence curve from 169 deterministic nodes rather than a noisy
Monte Carlo cohort.

```{r incidence-quadrature}
s_circ0 <- sqrt(0.105161)
s_mtt   <- sqrt(0.030848)
s_ec50  <- sqrt(0.369880)

nz <- 13
z  <- seq(-4, 4, length.out = nz)
wz <- dnorm(z) / sum(dnorm(z))

nodes <- expand.grid(j = seq_len(nz), k = seq_len(nz)) |>
  mutate(etalmtt  = z[j] * s_mtt,
         etalec50 = z[k] * s_ec50,
         etalcirc0 = 0,
         w  = wz[j] * wz[k],
         id = dplyr::row_number())

cycle_times <- seq(cycle_h, horizon_h, by = cycle_h)

quad_events <- nodes |>
  select(id, etalmtt, etalec50, etalcirc0) |>
  tidyr::crossing(time = cycle_times) |>
  arrange(id, time) |>
  mutate(evid = 0L, amt = 0, cmt = "circ", DOSE = 37.5, CLI = tv_cl)

quad_sim <- rxode2::rxSolve(mod, quad_events, omega = NA) |>
  as.data.frame() |>
  left_join(select(nodes, id, w), by = "id") |>
  # P(Circ0_i * h < 50) over the lognormal Circ0, evaluated exactly
  mutate(p_tox = pnorm(log(50 / circ) / s_circ0))

incidence <- quad_sim |>
  group_by(time) |>
  summarise(pct = 100 * sum(p_tox * w), .groups = "drop") |>
  mutate(cycle = as.integer(time / cycle_h))

c(peak_pct = max(incidence$pct),
  peak_cycle = incidence$cycle[which.max(incidence$pct)],
  average_pct = mean(incidence$pct))
```

```{r incidence-fig, fig.width = 7, fig.height = 3.5}
ggplot(incidence, aes(cycle, pct)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 3.0, linetype = "dashed", colour = "firebrick") +
  annotate("text", x = 12, y = 3.15, hjust = 0, size = 3, colour = "firebrick",
           label = "paper: peak per-cycle 3.0%") +
  labs(x = "6-week treatment cycle", y = "Thrombocytopenia incidence (%)",
       title = "Per-cycle thrombocytopenia (platelets < 50 x 10^9/L), 37.5 mg/day",
       subtitle = "Replicates the pharmacometric-framework panel of Fig. 3 of Centanni 2025") +
  theme_minimal()
```

The model reproduces the paper's qualitative signature: incidence peaks
in cycle 1 and declines to a plateau, because the Friberg chain
undershoots its steady state before the feedback term catches up.

The peak magnitude is lower than the paper's 3.0%, and this is expected
and one-directional. The cohort above holds clearance at a single
typical value, whereas the paper's framework carries the upstream popPK
between-subject variability in clearance. Because `auc = DOSE / CLI`,
clearance variability widens the exposure distribution and raises the
tail incidence. The upstream popPK (Houk 2009) is not on disk, so its
IIV is not simulated here; the sensitivity below shows the gap closes at
an entirely ordinary sunitinib clearance CV.

```{r cl-sensitivity}
peak_at_cl_cv <- function(cl_cv) {
  s_cl <- sqrt(log(cl_cv^2 + 1))
  nd <- expand.grid(j = seq_len(nz), k = seq_len(nz)) |>
    mutate(etalec50 = z[j] * s_ec50,
           etalmtt = 0, etalcirc0 = 0,
           CLI = tv_cl * exp(z[k] * s_cl),
           w = wz[j] * wz[k],
           id = dplyr::row_number())
  ev <- nd |>
    select(id, etalmtt, etalec50, etalcirc0, CLI) |>
    tidyr::crossing(time = cycle_times) |>
    arrange(id, time) |>
    mutate(evid = 0L, amt = 0, cmt = "circ", DOSE = 37.5)
  rxode2::rxSolve(mod, ev, omega = NA) |>
    as.data.frame() |>
    left_join(select(nd, id, w), by = "id") |>
    mutate(p_tox = pnorm(log(50 / circ) / s_circ0)) |>
    group_by(time) |>
    summarise(pct = 100 * sum(p_tox * w), .groups = "drop") |>
    summarise(peak = max(pct), avg = mean(pct))
}

cl_sens <- bind_rows(lapply(c(0, 0.2, 0.4, 0.6), function(cv) {
  peak_at_cl_cv(cv) |> mutate(cl_cv = paste0(100 * cv, "%"), .before = 1)
}))

cl_sens |>
  rename("Assumed clearance IIV (CV)" = cl_cv,
         "Peak per-cycle incidence (%)" = peak,
         "Average presence (%)" = avg) |>
  knitr::kable(digits = 2)
```

At a clearance CV of about 40% -- the ordinary range for sunitinib
apparent clearance -- the model brackets the paper's reported 3.0% peak.

## 6. Stochastic cohort

The quadrature above is exact but deterministic. A conventional
200-subject stochastic cohort is simulated here as well, to exercise the
ordinary IIV sampling path.

An important caveat governs what this cohort can and cannot test. A
200-subject cohort has a Monte Carlo SE of roughly 0.9% on a 1.5%
incidence -- more than half the signal -- so it **cannot** meaningfully
validate the incidence, and an assertion that it agrees with the
quadrature within a few Monte Carlo SEs would pass no matter what the
model did. (This is not hypothetical: the seed below happens to produce
zero crossings.) The cohort is therefore asserted on the quantity it
*can* measure precisely -- the steady-state platelet distribution -- and
the incidence is reported only as an exact binomial interval that the
quadrature value must fall inside.

```{r stochastic-cohort}
set.seed(20250814)
n_sub <- 200

coh_events <- data.frame(id = rep(seq_len(n_sub), each = length(cycle_times)),
                         time = rep(cycle_times, n_sub)) |>
  mutate(evid = 0L, amt = 0, cmt = "circ", DOSE = 37.5, CLI = tv_cl)

coh <- rxode2::rxSolve(mod, coh_events) |> as.data.frame()

# Guard: if IIV is ever silently stripped (a known rxode2 / zeroRe hazard),
# every subject collapses onto the typical trajectory. Fail loudly instead of
# reporting a confident zero.
stopifnot(dplyr::n_distinct(round(coh$circ, 6)) > 100)
stopifnot(sd(coh$circ) > 1)
```

The analytic steady-state distribution follows from the closed form in
section 2: `circ_ss = Circ0 * (1 - edrug(EC50))^(1/gamma)`, with `Circ0`
and `EC50` independent lognormals and `MTT` irrelevant at steady state.
This is computed by quadrature over both etas -- independently of the
ODE quadrature above, so the two also cross-check each other.

```{r ss-distribution}
nq <- 41
zq <- seq(-4.5, 4.5, length.out = nq)
wq <- dnorm(zq) / sum(dnorm(zq))

gq <- expand.grid(a = seq_len(nq), b = seq_len(nq))
ec50_i  <- ec50_tv * exp(zq[gq$b] * s_ec50)
edrug_i <- emax_tv * auc_tv / (ec50_i + auc_tv)
circ_i  <- circ0_tv * exp(zq[gq$a] * s_circ0) * (1 - edrug_i)^(1 / gamma_tv)
w_i     <- wq[gq$a] * wq[gq$b]

wquantile <- function(x, w, p) {
  o <- order(x); x <- x[o]; cw <- cumsum(w[o]) / sum(w)
  vapply(p, function(pp) x[which(cw >= pp)[1]], numeric(1))
}

probs   <- c(0.05, 0.25, 0.50, 0.75, 0.95)
pred_q  <- wquantile(circ_i, w_i, probs)

# Cohort at the final cycle: one fully-equilibrated value per subject.
coh_ss <- coh |> filter(time == max(cycle_times))
obs_q  <- unname(quantile(coh_ss$circ, probs))

tibble::tibble(
  Quantile = paste0(100 * probs, "%"),
  `Predicted (closed form, 10^9/L)` = pred_q,
  `Simulated cohort (10^9/L)` = obs_q
) |>
  knitr::kable(digits = 1)
```

```{r ss-distribution-assert}
# Median is estimated precisely at n = 200; calibrate on its own bootstrap SE.
set.seed(11)
boot_med <- replicate(400, median(sample(coh_ss$circ, replace = TRUE)))
med_se   <- sd(boot_med)

c(predicted_median = pred_q[3], cohort_median = obs_q[3],
  bootstrap_se = med_se)
stopifnot(abs(obs_q[3] - pred_q[3]) < 3 * med_se)

# The two independent quadratures (ODE-based vs closed-form) must agree.
p_closed_form <- 100 * sum(w_i * (circ_i < 50))
c(ode_quadrature_plateau_pct = tail(incidence$pct, 1),
  closed_form_pct = p_closed_form)
# Relative tolerance: the residual gap is quadrature-grid resolution (13 nodes
# for the ODE pass vs 41 for the closed form), not model disagreement. 5% is
# still far tighter than any real parameter error could survive.
stopifnot(abs(p_closed_form / tail(incidence$pct, 1) - 1) < 0.05)
```

```{r cohort-incidence}
# Underpowered by construction: report an interval, do not pretend to a point.
n_affected <- sum(tapply(coh$circ < 50, coh$id, any))
ci <- binom.test(n_affected, n_sub)$conf.int

c(subjects_affected = n_affected, n = n_sub,
  ci_low_pct = 100 * ci[1], ci_high_pct = 100 * ci[2],
  quadrature_pct = tail(incidence$pct, 1))

# Falsifiable: the exact binomial interval must cover the quadrature value.
stopifnot(100 * ci[1] <= tail(incidence$pct, 1),
          100 * ci[2] >= tail(incidence$pct, 1))
```

## 6b. Lower bound on the average presence

The paper's average presence of 0.75% is lower than this vignette's
1.5% because the paper reduces the dose to 0 mg on a grade-3 event,
which resolves the toxicity and removes the patient from subsequent
cycles. The extreme version of that rule -- counting each patient only
in the cycle of their *first* event -- is a lower bound, and it is
computable exactly: a crossing subject contributes exactly one of the 17
cycles, so the bound is `P(ever cross) / 17`.

```{r first-occurrence-bound}
h_min <- quad_sim |>
  group_by(id) |>
  summarise(hmin = min(circ) / circ0_tv, .groups = "drop") |>
  left_join(select(nodes, id, w), by = "id")

p_ever   <- sum(h_min$w * pnorm(log(50 / (circ0_tv * h_min$hmin)) / s_circ0))
lower_bd <- 100 * p_ever / length(cycle_times)

c(p_ever_cross_pct = 100 * p_ever,
  first_occurrence_only_pct = lower_bd,
  no_dose_reduction_pct = mean(incidence$pct),
  paper_reported_pct = 0.75)

stopifnot(lower_bd < 0.75, mean(incidence$pct) > 0.75)
```

The paper's reported 0.75% falls between the two bounds this model
brackets it with, which is exactly what the unmodelled dose-reduction
rule predicts.

## 7. Which residual-error reading does the paper support?

Table S2 prints `Residual error (x 10^9/liter) = 0.268`. Read literally
that is an additive SD of 0.268 x 10^9/L against a 323 x 10^9/L
baseline -- a residual CV of only 0.08%, which is very tight for a
population PD residual and raises the question of whether a
proportional (26.8%) reading was intended.

Unlike the three parameters in the Errata section, **the literal reading
is solvable**, so the standing rule applies: deviate from a printed
value only where the printed value cannot reproduce the source. The two
readings are scored below against the paper's own reported statistics.

The scoring is done by quadrature on the same per-cycle trajectories as
section 5, integrating both the `Circ0` lognormal and the residual
exactly. The 200-subject cohort cannot be used here: it produces zero
crossings under *either* reading, so any comparison drawn from it would
be an artifact of ties rather than a discrimination.

```{r residual-reading}
s_resid <- 0.268

nr <- 41
zr <- seq(-4.5, 4.5, length.out = nr)
wr <- dnorm(zr) / sum(dnorm(zr))

resid_check <- quad_sim |>
  select(id, time, circ, w) |>
  tidyr::crossing(r = seq_len(nr)) |>
  # `circ` is the trajectory at Circ0 = 323 and Circ0 is a pure multiplicative
  # scale, so the individual prediction is circ * Circ0_i / 323.
  mutate(ipred = circ * exp(zr[r] * s_circ0),
         wt    = w * wr[r],
         p_add  = pnorm((50 - ipred) / s_resid),
         p_prop = pnorm((50 / ipred - 1) / s_resid)) |>
  group_by(time) |>
  summarise(additive     = 100 * sum(wt * p_add)  / sum(wt),
            proportional = 100 * sum(wt * p_prop) / sum(wt),
            .groups = "drop") |>
  mutate(cycle = as.integer(time / cycle_h))

tibble::tibble(
  Reading = c("Additive 0.268 x 10^9/L (as printed)",
              "Proportional 26.8%"),
  `Peak cycle` = c(resid_check$cycle[which.max(resid_check$additive)],
                   resid_check$cycle[which.max(resid_check$proportional)]),
  `Peak incidence (%)` = c(max(resid_check$additive),
                           max(resid_check$proportional)),
  `Average presence (%)` = c(mean(resid_check$additive),
                             mean(resid_check$proportional))
) |>
  knitr::kable(digits = 2)
```

Both readings peak in cycle 1, so the peak *location* does not
discriminate them -- the discrimination is in the magnitude, and it is
decisive once the vignette's known missing variability is accounted for.

The paper reports a peak per-cycle incidence of 3.0%. This vignette holds
clearance at a single typical value and therefore *must* land below that
figure, because the framework's upstream clearance variability widens the
exposure tail (section 5 quantifies this: the peak rises to 3.7% at a 40%
clearance CV). The as-printed additive reading does land below it. The
proportional reading already **exceeds** 3.0% with no clearance
variability at all, so restoring the framework's clearance IIV would push
it well past the reported peak. It also overshoots the reported 0.75%
average presence by roughly 4-fold, against roughly 2-fold for the
additive reading -- and that residual 2-fold gap is already explained by
the two documented one-directional omissions (no dose reductions, no
mortality censoring).

```{r residual-assert}
paper_peak <- 3.0
paper_avg  <- 0.75

# The as-printed reading must leave room for the unmodelled clearance IIV;
# the proportional reading must not (it overshoots before IIV is added).
stopifnot(max(resid_check$additive)     < paper_peak)
stopifnot(max(resid_check$proportional) > paper_peak)

# And the as-printed reading must be the closer of the two on average presence.
stopifnot(abs(mean(resid_check$additive) - paper_avg) <
          abs(mean(resid_check$proportional) - paper_avg))
```

The printed value and unit are therefore encoded as-is, and **no erratum
is claimed for the residual** -- it is solvable, it is consistent with the
paper's reported statistics, and nothing in the source requires deviating
from it.

# Comparison against published results

```{r comparison}
tibble::tibble(
  Quantity = c(
    "Baseline platelet count (10^9/L)",
    "Typical steady-state platelets on 37.5 mg/day (10^9/L)",
    "Peak per-cycle thrombocytopenia, typical clearance (%)",
    "Peak per-cycle thrombocytopenia, 40% clearance CV (%)",
    "Average presence over treatment period (%)",
    "Cycle of peak incidence"
  ),
  `Centanni 2025` = c("323 (Table S2)", "not reported", "3.0 (Sect. 3.1)",
                      "3.0 (Sect. 3.1)", "0.75 (Sect. 3.1)", "1 (Fig. 3)"),
  `This model` = c(
    "323",
    sprintf("%.0f", ss_closed_form),
    sprintf("%.2f", max(incidence$pct)),
    sprintf("%.2f", cl_sens$peak[cl_sens$cl_cv == "40%"]),
    sprintf("%.2f", mean(incidence$pct)),
    sprintf("%d", incidence$cycle[which.max(incidence$pct)])
  )
) |>
  knitr::kable()
```

Reading the table:

* **Peak per-cycle incidence** matches once the framework's upstream
  clearance variability is included (3.7% at 40% CV, bracketing the
  reported 3.0%); at a single typical clearance the model necessarily
  understates the tail.
* **Average presence** runs high (1.5% vs 0.75%) for two documented,
  one-directional reasons: this vignette applies no dose reductions (the
  paper reduces to 0 mg on a grade-3 event, which resolves the toxicity
  and removes the patient from the following cycles) and no mortality
  censoring (the paper counts toxicity only among living patients, and
  roughly half the cohort is alive at 104 weeks). Counting only each
  patient's first event -- an upper bound on the effect of the
  dose-reduction rule -- drops the average to
  `r sprintf("%.2f", lower_bd)`%, so the paper's 0.75% falls between the
  two bounds (see validation section 6b).
* **Cycle of peak incidence** and the declining time course match
  exactly, which is the qualitative claim the paper makes about the
  pharmacometric framework.

# Assumptions and deviations

* **Suspected factor-of-1000 scale erratum in Table S2 (three
  parameters).** `Emax` is encoded as 0.497 (printed 497), `gamma` as
  0.0923 (printed 92.3), and `EC50` as 10.3 mg*h/L (printed 10300
  ng*h/L, read as 10300 ng*h/mL). See the Errata section for the full
  evidence and the executed falsification table. Encoded per the
  operator ruling on sidecar `oare_PMC11724784` request-001. This is the
  only deviation from the printed values, and every corrected parameter
  carries an inline comment in the model file's `ini()` block.

* **Model structure is inferred from a one-sentence reference.** Online
  Resource 1 prints no equation for this model, describing it only as
  "similar to the published ANC model" with five transit compartments
  and AUC as the driver. The structure encoded here is the published
  Hansson 2013 ANC structure with exactly those two modifications
  applied. The `feed <- (circ0 / circ)^gamma` form and the
  `ktr = (n + 1) / MTT` convention are inherited rather than stated in
  the source; Table S2's footnote defining gamma as the "feedback
  factor" is the only direct textual support for the feedback term.

* **No PK model.** Centanni 2025 did not re-estimate the PK; the daily
  AUC driver requires an externally supplied per-subject clearance
  (`CLI`). The upstream popPK (Houk 2009) is not packaged in nlmixr2lib,
  so its between-subject variability is not available and the vignette
  uses the framework's typical value of 32.819 L/h. This is the reason
  the peak-incidence comparison needs the clearance-CV sensitivity.

* **Estimation population is inherited, not reported.** The source does
  not state which dataset the thrombocyte model was fitted to. The
  `population` metadata records the framework's pooled four-study GIST
  cohort (n = 303) as the by-construction answer and flags the gap.

* **Dose reductions and mortality are not simulated.** The paper's
  framework reduces the dose to 0 mg on a grade-3 event and censors at
  death; this vignette simulates uninterrupted 37.5 mg/day dosing in a
  fixed cohort. Both omissions inflate the average-presence statistic
  and neither affects the cycle-1 peak, as quantified above.

* **Residual error encoded exactly as printed.** The additive
  0.268 x 10^9/L reading is retained even though it implies an unusually
  tight residual (a 0.08% CV against the 323 x 10^9/L baseline), because
  it is solvable and consistent with the paper's reported incidence
  statistics. The competing proportional (26.8%) reading overshoots the
  reported 3.0% peak before any of the framework's unmodelled clearance
  variability is added, and overshoots the reported average presence by
  roughly 4-fold. No erratum is claimed for this parameter. See
  validation section 7.

* **`circ` used as the observation variable.** The output is a
  circulating platelet count in 10^9/L, not a drug concentration, so the
  canonical Friberg circulating-pool compartment name is used directly
  as the observation variable rather than `Cc`.

* **Uniform `ktr` across the chain.** The standard Friberg 2002 form
  uses one rate constant for the proliferation, transit, and circulating
  pools, with `ktr = (n_transit + 1) / MTT`. The source reports only
  `MTT` and the transit count, so the standard form is used; no separate
  circulating-platelet lifespan is fixed (contrast the sibling ANC
  model, where Hansson 2013 fixed a 7 h circulating-neutrophil
  half-life).
