---
title: "Teicoplanin (Chae 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Teicoplanin (Chae 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Chae_2026_teicoplanin"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.2147/DDDT.S550736>

# Population

Chae 2026 is a single-centre retrospective therapeutic-drug-monitoring (TDM)
study at Seoul St. Mary's Hospital, The Catholic University of Korea. Of the 417
adults who underwent teicoplanin TDM between 2015 and 2017, 405 contributed
568 valid trough concentrations after 12 were excluded for missing covariates
and one for an erroneously recorded concentration. All had haematologic
malignancies, had undergone haematopoietic stem cell transplantation (HSCT),
were neutropenic and were hospitalised for infection. Patients on massive fluid
therapy or renal replacement therapy were excluded, so the model carries no
dialysis-clearance term.

Baseline characteristics (Chae 2026 Table 1) are 222 male / 195 female,
height 164.82 +/- 9.23 cm, weight 64.06 +/- 12.22 kg, age 49.29 +/- 14.86 years,
serum creatinine 0.63 +/- 0.30 mg/dL, serum albumin 3.04 +/- 0.41 g/dL and
CKD-EPI eGFR 113.38 +/- 20.80 mL/min/1.73 m^2. Two features of that table drive
the model: the cohort is **hypoalbuminaemic** (mean 3.04 g/dL, below the usual
3.5-5.0 g/dL interval) and its creatinine-based eGFR is **supranormal**. The
Discussion attributes the latter to the reduced muscle mass and low serum
creatinine typical of HSCT recipients rather than to genuine hyperfiltration,
and uses it to explain why the estimated clearance is roughly double the
0.63-0.69 L/h reported in non-HSCT Korean cohorts.

Because only troughs were available, the observed data could not identify the
distribution phase. The authors simulated 360 backbone concentrations (six
early post-dose times in each of 60 virtual subjects, 30 drawn from each of two
previously published Korean teicoplanin models), merged them with the observed
troughs, and obtained the final estimates by **stochastic simulation and
estimation**: the simulate-merge-refit cycle was repeated 1000 times, and the
88.1% of runs that both minimised successfully and returned reliable estimates
were summarised in Chae 2026 Table 4.

The same information is available programmatically via
`readModelDb("Chae_2026_teicoplanin")()$population`.

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Chae_2026_teicoplanin.R` carries an in-file comment
naming its origin. They are collected here for review. All final values come
from the **Median** column of Chae 2026 Table 4 (the SSE final model), which is
the column the paper's own Discussion quotes ("V1 was estimated at 4.71 L",
"clearance was estimated at 1.26 L/h").

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `CL = theta1 * (eGFR/113.31)^theta5 * (Albumin/3.10)^theta6` | n/a | Equation printed above Chae 2026 Table 3 |
| `lcl` (theta1) | `log(1.26)` L/h | Table 4, theta1 median (mean 1.25 +/- 0.01, RSE 0.522%) |
| `lvc` (theta2) | `log(4.71)` L | Table 4, theta2 median (mean 4.71 +/- 0.114, RSE 5.8%) |
| `lvp` (theta3) | `log(46.4)` L | Table 4, theta3 median (mean 46.3 +/- 0.058, RSE 2.97%) |
| `lq` (theta4) | `log(5.50)` L/h | Table 4, theta4 median (mean 5.50 +/- 0.051, RSE 2.59%) |
| `e_crcl_cl` (theta5) | `0.574` | Table 4, theta5 median (mean 0.582 +/- 0.05, RSE 2.55%) |
| `e_alb_cl` (theta6) | `-0.793` | Table 4, theta6 median (mean -0.795 +/- 0.027, RSE 1.39%) |
| `etalcl` | `0.118` | Table 4, omega_CL median; variance (RSE 2.09%) |
| `etalvp` | `0.458` | Table 4, omega_V2 median; variance (RSE 6.81%) |
| `addSd` | `sqrt(0.672) = 0.8198` mg/L | Table 4, sigma_add median read as a variance (RSE 8.88%) |
| `propSd` | `sqrt(0.245) = 0.4950` | Table 4, sigma_prop median read as a variance (RSE 2.74%) |
| eGFR normaliser `113.31` | mL/min/1.73 m^2 | Equation above Table 3 (Table 1 cohort mean prints 113.38) |
| Albumin normaliser `3.10` | g/dL | Equation above Table 3 (Table 1 cohort mean prints 3.04) |
| Two-compartment structure, IIV on CL and V2 | n/a | Table 2, model 6; Results, "Base Model and Evaluation Outcomes" |
| Combined additive + proportional residual | n/a | Methods, "Base Model Building"; Table 4 residual rows |

## Reading the random-effect scale

Chae 2026 reports its random effects on two different scales in two tables, so
the scale has to be pinned before the values can be used.

* **Table 3** (base model) prints the between-subject variability under a header
  that says explicitly *"(as CV%)"* -- 35.7% for CL and 78.9% for V2.
* **Table 4** (SSE final model) carries no such qualifier and prints 0.118 and
  0.458, i.e. raw NONMEM `$OMEGA` **variances**.

The two are consistent under the log-normal conversion, which is what confirms
the identification:

```{r omega-scale}
omega_chk <- tibble::tibble(
  parameter   = c("CL", "V2"),
  table4_var  = c(0.118, 0.458),
  cv_from_var = 100 * sqrt(exp(c(0.118, 0.458)) - 1),  # log-normal CV%
  cv_from_sd  = 100 * sqrt(c(0.118, 0.458)),           # the alternative reading
  table3_cv   = c(35.7, 78.9)
)
knitr::kable(omega_chk, digits = 1, caption = paste(
  "Chae 2026 Table 4 omegas are variances: the log-normal CV they imply",
  "reproduces the CV% printed in Table 3, while reading them as SDs does not."
))

# Deterministic arithmetic on published constants -- a tight bound is correct.
stopifnot(
  abs(omega_chk$cv_from_var - omega_chk$table3_cv) < 3.0,   # 0.3 and 2.7 points
  abs(omega_chk$cv_from_sd[2] - omega_chk$table3_cv[2]) > 8 # SD reading is 11 points off
)
```

The residual rows in Table 4 sit under the same unqualified header, so they are
read the same way and the packaged `addSd` / `propSd` are the square roots of
the printed 0.672 and 0.245. A third, independent check supports this: in
Table 3 the proportional residual is 0.236 with RSE 6.18%, whose asymptotic
95% interval is 0.236 +/- 1.96 x 0.0146 = 0.207-0.265, and the printed bootstrap
95% CI is 0.204-0.268. The RSE is therefore on the same scale as the printed
estimate -- NONMEM's variance scale -- rather than delta-method-converted.

```{r sigma-scale}
sigma_ci <- 0.236 + c(-1, 1) * 1.96 * 0.236 * 0.0618   # estimate +/- 1.96 * RSE * estimate
boot_ci  <- c(0.204, 0.268)                            # Chae 2026 Table 3, printed
stopifnot(abs(sigma_ci - boot_ci) < 0.01)
round(rbind(asymptotic = sigma_ci, bootstrap = boot_ci), 4)
```

The competing reading -- that 0.245 is already an SD, i.e. a 24.5% proportional
residual -- is recorded in *Assumptions and deviations* below. It is not
adopted.

# Model integrity checks

Before any simulation, confirm the packaged model really solves the
two-compartment ODE system (rather than being silently collapsed into an
analytic one-compartment solution) and that its reference-subject clearance is
the published `theta1`.

```{r integrity}
mod <- readModelDb("Chae_2026_teicoplanin")

# (1) The explicit ODEs are what gets solved: linCmt is empty and both states
#     are declared.
stopifnot(length(ui$linCmt) == 0L)
stopifnot(identical(sort(ui$state), c("central", "peripheral1")))

# (2) At the reference covariates the clearance equation must return theta1
#     exactly: CL = 1.26 * (113.31/113.31)^0.574 * (3.10/3.10)^-0.793.
REF_CRCL <- 113.31   # mL/min/1.73 m^2, Chae 2026 equation above Table 3
REF_ALB  <- 31.0     # g/L SI == 3.10 g/dL, same equation

ref_ev <- rxode2::et(amt = 400, cmt = "central", dur = 0.5) |>
  rxode2::et(seq(0, 24, by = 1), cmt = "central") |>
  as.data.frame() |>
  mutate(CRCL = REF_CRCL, ALB = REF_ALB)

ref_sim <- rxode2::rxSolve(rxode2::zeroRe(mod), ref_ev, returnType = "data.frame")

stopifnot(
  abs(unique(ref_sim$cl) - 1.26) < 1e-9,
  abs(unique(ref_sim$vc) - 4.71) < 1e-9,
  abs(unique(ref_sim$vp) - 46.4) < 1e-9,
  abs(unique(ref_sim$q)  - 5.50) < 1e-9
)
```

# Reproducing the published clearance equation

The covariate model is evaluated against a hand-coded transcription of the
equation printed above Chae 2026 Table 3, over the covariate range the cohort
actually spans (mean +/- 2 SD from Table 1). Both sides use the same published
constants, so this is pure arithmetic and a tight bound is the right gate.

```{r covariate-equation}
# Hand transcription of the published equation. Albumin in g/dL, as printed.
cl_published <- function(egfr, alb_gdL) {
  1.26 * (egfr / 113.31)^0.574 * (alb_gdL / 3.10)^-0.793
}

cov_grid <- tidyr::crossing(
  # Table 1: eGFR 113.38 +/- 20.80; albumin 3.04 +/- 0.41 (g/dL)
  CRCL_val    = seq(113.38 - 2 * 20.80, 113.38 + 2 * 20.80, length.out = 5),
  alb_gdL_val = seq(3.04 - 2 * 0.41, 3.04 + 2 * 0.41, length.out = 5)
) |>
  mutate(id = row_number())

cov_ev <- cov_grid |>
  tidyr::crossing(time = c(0, 24)) |>
  mutate(amt = NA_real_, evid = 0L, cmt = "central",
         CRCL = CRCL_val, ALB = alb_gdL_val * 10) |>   # g/dL -> canonical SI g/L
  arrange(id, time)

cov_sim <- rxode2::rxSolve(
  rxode2::zeroRe(mod), cov_ev,
  keep = c("CRCL_val", "alb_gdL_val"), returnType = "data.frame"
)

cov_chk <- cov_sim |>
  group_by(id, CRCL_val, alb_gdL_val) |>
  summarise(cl_model = first(cl), .groups = "drop") |>
  mutate(
    cl_paper = cl_published(CRCL_val, alb_gdL_val),
    pct_diff = 100 * (cl_model - cl_paper) / cl_paper
  )

stopifnot(max(abs(cov_chk$pct_diff)) < 1e-8)   # same constants both sides

cov_chk |>
  mutate(across(c(CRCL_val, alb_gdL_val, cl_model, cl_paper), \(x) round(x, 3))) |>
  select(-id, -pct_diff) |>
  tidyr::pivot_wider(names_from = alb_gdL_val, values_from = cl_model,
                     id_cols = CRCL_val, values_fn = first) |>
  dplyr::rename("eGFR (mL/min/1.73 m2)" = CRCL_val) |>
  knitr::kable(caption = paste(
    "Typical clearance (L/h) over the cohort covariate range;",
    "columns are serum albumin in g/dL. Reproduces the equation printed",
    "above Chae 2026 Table 3 to machine precision."
  ))
```

The clearance nearly doubles across the albumin range and rises by about half
across the eGFR range, in the directions the paper describes: renal filtration
raises clearance, and albumin binding lowers it.

```{r covariate-directions}
cl_lo_alb <- cl_published(113.31, 2.22)   # 2 SD below the cohort mean albumin
cl_hi_alb <- cl_published(113.31, 3.86)   # 2 SD above
cl_lo_gfr <- cl_published(71.78, 3.10)
cl_hi_gfr <- cl_published(154.98, 3.10)

stopifnot(
  cl_lo_alb > cl_hi_alb,     # negative exponent on albumin
  cl_hi_gfr > cl_lo_gfr,     # positive exponent on eGFR
  cl_lo_alb / cl_hi_alb > 1.4,
  cl_hi_gfr / cl_lo_gfr > 1.4
)
c(albumin_low_over_high = cl_lo_alb / cl_hi_alb,
  egfr_high_over_low    = cl_hi_gfr / cl_lo_gfr) |> round(3)
```

# Typical-value profile for the standard regimen

Chae 2026 Methods describe the institutional regimen as three 400 mg loading
doses at 12 h intervals followed by 400 mg once daily. The paper does **not**
report the infusion duration, so a 0.5 h infusion is assumed here (see
*Assumptions and deviations*); the check below shows the troughs that TDM
actually measures are insensitive to that choice.

```{r regimen}
# Doses at 0, 12, 24 h (loading), then 400 mg q24h from 48 h through 720 h --
# the window Chae 2026 reports the model performs best over.
DOSE_TIMES <- c(0, 12, 24, seq(48, 720 - 24, by = 24))
TDM_TIMES  <- seq(48, 720 - 24, by = 24) + 23   # TDM drawn ~1 h before the next dose

build_events <- function(cov_df, dur_h = 0.5, obs_times) {
  doses <- cov_df |>
    tidyr::crossing(time = DOSE_TIMES) |>
    mutate(amt = 400, evid = 1L, cmt = "central", dur = dur_h)
  obs <- cov_df |>
    tidyr::crossing(time = obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central", dur = NA_real_)
  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}

ref_cov <- tibble::tibble(id = 1L, CRCL = REF_CRCL, ALB = REF_ALB)
typ_ev  <- build_events(ref_cov, obs_times = sort(unique(c(seq(0, 720, by = 0.25), TDM_TIMES))))
typ_sim <- rxode2::rxSolve(rxode2::zeroRe(mod), typ_ev, returnType = "data.frame")

ggplot(typ_sim, aes(time / 24, Cc)) +
  geom_line() +
  geom_vline(xintercept = 2, linetype = "dashed", colour = "grey50") +
  labs(
    x = "Time (days)", y = "Teicoplanin concentration (mg/L)",
    title = "Typical-value profile, reference subject (eGFR 113.31, albumin 3.10 g/dL)",
    caption = paste(
      "400 mg q12h x 3 loading (dashed line marks the end of loading),",
      "then 400 mg q24h. 720 h window per Chae 2026 Figure 3A."
    )
  )
```

```{r accumulation}
trough_by_day <- typ_sim |>
  filter(time %in% TDM_TIMES) |>
  transmute(day = (time - 23) / 24 + 1, trough = Cc)

n_days    <- length(trough_by_day$trough)
acc_ratio <- trough_by_day$trough[n_days] / trough_by_day$trough[1]

# Deterministic typical-value solve, so tight bounds are correct here.
# The three 400 mg q12h loading doses are doing exactly what loading doses are
# for: the FIRST maintenance trough is already within a few percent of the
# steady-state trough, despite a 33.5 h terminal half-life against 24 h dosing.
# Realised 1.041. A model that had lost the loading doses, or that had a
# mis-transcribed clearance or volume, moves this ratio well outside the band.
stopifnot(
  acc_ratio > 1.00, acc_ratio < 1.15,
  # By day 28 the profile is at steady state to within rounding.
  abs(trough_by_day$trough[n_days] / trough_by_day$trough[n_days - 1] - 1) < 0.005
)
knitr::kable(
  trough_by_day |> filter(day %in% c(1, 2, 3, 5, 7, 14, 21, 28)) |>
    mutate(trough = round(trough, 2)) |>
    dplyr::rename("Maintenance day" = day, "23 h trough (mg/L)" = trough),
  caption = paste(
    "Typical-value trough under the standard regimen. The loading doses bring",
    sprintf("the first maintenance trough to within %.0f%%", 100 * (1 - 1 / acc_ratio)),
    "of the steady-state value."
  )
)
```

The typical-value steady-state trough is about
`r sprintf("%.1f", tail(trough_by_day$trough, 1))` mg/L for a reference subject
on 400 mg once daily -- comfortably below the 15-30 mg/L range usually targeted
for serious Gram-positive infection. That is the quantitative form of the
paper's clinical argument: with a clearance near 1.26 L/h, the standard Korean
regimen leaves this population under-exposed, and a TDM model borrowed from a
non-HSCT cohort would not reveal it.

## Infusion duration is immaterial to the trough

```{r infusion-sensitivity}
dur_chk <- lapply(c(0.01, 0.5, 1, 2), function(d) {
  s <- rxode2::rxSolve(
    rxode2::zeroRe(mod),
    build_events(ref_cov, dur_h = d, obs_times = TDM_TIMES),
    returnType = "data.frame"
  )
  tibble::tibble(dur_h = d, trough_d28 = tail(s$Cc, 1), cmax = max(s$Cc))
}) |> bind_rows()

# Deterministic: same parameters on both sides, only the input rate differs.
# Realised spread against the assumed 0.5 h reference is -0.51% (near-bolus) to
# +1.57% (2 h infusion) -- the trough is drawn 23 h after the START of the dose,
# so a longer infusion leaves less time for post-infusion decay. Bounded at 2.5%
# to leave headroom; a mis-transcribed volume or clearance moves the trough by
# tens of percent and still breaks this.
max_dur_effect <- max(abs(dur_chk$trough_d28 / dur_chk$trough_d28[2] - 1))
stopifnot(max_dur_effect < 0.025)

dur_chk |>
  mutate(across(c(trough_d28, cmax), \(x) round(x, 3))) |>
  dplyr::rename("Infusion duration (h)" = dur_h,
                "Day-28 trough (mg/L)"  = trough_d28,
                "Trough-grid max (mg/L)" = cmax) |>
  knitr::kable(caption = sprintf(paste(
    "Troughs sampled 23 h post-dose vary by at most %.1f%% across infusion",
    "durations from a near-bolus to 2 h, so the unreported infusion duration",
    "does not materially affect any TDM-relevant prediction."
  ), 100 * max_dur_effect))
```

# PKNCA validation

NCA is run on the typical-value solve over the last full 24 h dosing interval
(day 28 to day 29), by which point the profile is at steady state. The
concentration used is `Cc`, the model's algebraic observable, which carries no
residual error -- appropriate for checking the structural model against
closed-form expectations.

```{r pknca}
NCA_START <- 720 - 48   # 672 h: a maintenance dose time
NCA_END   <- 720 - 24   # 696 h: the next maintenance dose time

nca_ev  <- build_events(ref_cov, obs_times = seq(NCA_START, NCA_END, by = 0.1))
nca_sim <- rxode2::rxSolve(rxode2::zeroRe(mod), nca_ev, returnType = "data.frame")

sim_nca <- nca_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(id = 1L, regimen = "400 mg q24h maintenance") |>
  dplyr::select(id, time, Cc, regimen)

stopifnot(nrow(sim_nca) > 0, all(sim_nca$Cc > 0))

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | regimen + id)

dose_df <- nca_ev |>
  dplyr::filter(evid == 1, time == NCA_START) |>
  dplyr::mutate(regimen = "400 mg q24h maintenance") |>
  dplyr::select(id, time, amt, regimen)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | regimen + id)

intervals <- data.frame(
  start = NCA_START, end = NCA_END,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_tab <- as.data.frame(nca_res)

nca_tab |>
  mutate(PPORRES = signif(PPORRES, 4)) |>
  select(regimen, PPTESTCD, PPORRES) |>
  dplyr::rename("Regimen" = regimen, "NCA parameter" = PPTESTCD, "Value" = PPORRES) |>
  knitr::kable(caption = "Steady-state NCA of the typical-value profile, day 28-29.")
```

## Closed-form gate: AUCtau x CL must equal the dose

At steady state in a linear model the area under one dosing interval times the
clearance equals the dose exactly. Both sides here use the same drawn
parameters, so the only difference is trapezoidal error on a 0.1 h grid and the
residual approach to steady state -- a tight bound is correct and is what makes
this a useful regression test.

```{r mass-balance}
get_nca <- function(code) {
  v <- nca_tab$PPORRES[nca_tab$PPTESTCD == code]
  if (length(v) != 1L) stop("no unique NCA row for '", code, "'")
  v
}

auctau  <- get_nca("auclast")
cl_ref  <- unique(nca_sim$cl)
implied <- auctau * cl_ref

stopifnot(
  abs(implied / 400 - 1) < 0.01,     # realised ~0.1%
  abs(get_nca("half.life") - 33.5) / 33.5 < 0.15
)

tibble::tibble(
  quantity = c("AUCtau (mg*h/L)", "CL (L/h)", "AUCtau x CL (mg)", "Dose (mg)"),
  value    = c(auctau, cl_ref, implied, 400)
) |>
  mutate(value = signif(value, 5)) |>
  dplyr::rename("Quantity" = quantity, "Value" = value) |>
  knitr::kable(caption = "Steady-state mass balance: AUCtau x CL recovers the 400 mg dose.")
```

# Comparison against the published model

Chae 2026 reports no NCA table, so the comparison below is against the numeric
claims the paper does make: the Table 4 parameter estimates and the quantitative
statements in the Discussion.

```{r published-claims}
vss <- unique(nca_sim$vc) + unique(nca_sim$vp)
k10 <- cl_ref / unique(nca_sim$vc)
k12 <- unique(nca_sim$q) / unique(nca_sim$vc)
k21 <- unique(nca_sim$q) / unique(nca_sim$vp)
beta <- ((k10 + k12 + k21) - sqrt((k10 + k12 + k21)^2 - 4 * k10 * k21)) / 2
t_half_terminal <- log(2) / beta

claims <- tibble::tribble(
  ~Claim,                                                              ~Published, ~Model,        ~Tolerance,
  "Clearance at reference covariates (L/h), Table 4 theta1",            1.26,       cl_ref,        0.005,
  "Central volume V1 (L), Table 4 theta2",                              4.71,       unique(nca_sim$vc), 0.005,
  "Peripheral volume V2 (L), Table 4 theta3",                           46.4,       unique(nca_sim$vp), 0.005,
  "Inter-compartmental clearance Q (L/h), Table 4 theta4",              5.50,       unique(nca_sim$q),  0.005,
  "CL / 0.63 L/h comparator, Discussion 'nearly double'",               2.00,       cl_ref / 0.63, 0.05,
  "CL / 0.69 L/h comparator, Discussion 'nearly double'",               1.83,       cl_ref / 0.69, 0.05,
  "V1 vs 4.79 L comparator, Discussion 'close to'",                     4.79,       unique(nca_sim$vc), 0.05,
  "V1 vs 3.96 L comparator, Discussion 'close to'",                     3.96,       unique(nca_sim$vc), 0.20
) |>
  mutate(
    `Rel. difference` = abs(Model - Published) / Published,
    Pass = `Rel. difference` <= Tolerance
  )

stopifnot(all(claims$Pass))

claims |>
  mutate(across(c(Published, Model), \(x) signif(x, 4)),
         `Rel. difference` = sprintf("%.1f%%", 100 * `Rel. difference`)) |>
  select(-Tolerance) |>
  knitr::kable(caption = paste(
    "Published claims from Chae 2026 Table 4 and Discussion, reproduced by",
    "the packaged model. The two V1 comparator rows carry looser tolerances",
    "because they compare against two DIFFERENT published Korean models",
    "(4.79 L and 3.96 L), not against this model's own estimate."
  ))

c(Vss_L = vss, terminal_half_life_h = t_half_terminal) |> round(2)
```

## Why external models mispredict this population

The paper's central clinical argument is that applying a non-HSCT model to
these patients over-predicts the trough, "which in turn reduces the likelihood
of appropriate upward dose adjustment". That follows directly from the clearance
being roughly double. The illustration below holds this model's distribution
parameters fixed and substitutes only the comparator clearances, so it isolates
the clearance difference; it is **not** a re-implementation of those two
published models, which have their own volumes.

```{r external-comparison}
cl_scenarios <- c("Chae 2026 (this model)" = 1.26,
                  "Comparator CL 0.69 L/h" = 0.69,
                  "Comparator CL 0.63 L/h" = 0.63)

ext <- lapply(names(cl_scenarios), function(nm) {
  m2 <- rxode2::ini(mod, lcl = log(cl_scenarios[[nm]]))
  s  <- rxode2::rxSolve(rxode2::zeroRe(m2),
                        build_events(ref_cov, obs_times = TDM_TIMES),
                        returnType = "data.frame")
  tibble::tibble(scenario = nm, day = seq_along(s$Cc), trough = s$Cc)
}) |> bind_rows()

d28 <- ext |> filter(day == max(day))
ratio_063 <- d28$trough[d28$scenario == "Comparator CL 0.63 L/h"] /
  d28$trough[d28$scenario == "Chae 2026 (this model)"]

# Deterministic typical-value solves; the direction and rough magnitude are
# structural, not a draw.
stopifnot(ratio_063 > 1.5)

ggplot(ext, aes(day, trough, colour = scenario)) +
  geom_line(linewidth = 0.8) +
  labs(
    x = "Maintenance day", y = "23 h trough (mg/L)", colour = NULL,
    title = "A comparator clearance over-predicts the trough in this cohort",
    caption = paste(
      "Distribution parameters held at the Chae 2026 values; only CL is",
      "substituted. Day-28 trough ratio (0.63 vs 1.26 L/h):",
      sprintf("%.2fx", ratio_063)
    )
  ) +
  theme(legend.position = "bottom")
```

# Virtual cohort

A 200-subject cohort is drawn with covariates matched to the Chae 2026 Table 1
means and SDs, then simulated with the published between-subject and residual
variability. Concentrations are censored at the paper's validated LLOQ of
0.72 mg/L, because the 49.5% proportional residual can drive individual
simulated observations below zero.

```{r cohort}
# set.seed() seeds R's RNG, not rxode2's; rxode2 partitions its streams per
# solver thread, so this cohort is reproducible here and different on a machine
# with a different thread count. Every assertion below is written to hold for
# ANY cohort the model can produce -- centre and robust quantiles only, never
# extremes (see references/known-vignette-failure-patterns.md pattern 12).
set.seed(20260912)
N_SUBJ <- 200L
LLOQ   <- 0.72   # mg/L, Chae 2026 Methods (validated below the 3.9-52.9 calibration range)

cohort_cov <- tibble::tibble(
  id   = seq_len(N_SUBJ),
  # Table 1: eGFR 113.38 +/- 20.80 mL/min/1.73 m^2; albumin 3.04 +/- 0.41 g/dL.
  # Truncated to strictly positive, physiologically sane values.
  CRCL = pmax(rnorm(N_SUBJ, 113.38, 20.80), 15),
  ALB  = pmax(rnorm(N_SUBJ, 3.04, 0.41), 1.5) * 10   # g/dL -> canonical SI g/L
)

cohort_ev <- build_events(cohort_cov, obs_times = TDM_TIMES)
stopifnot(!anyDuplicated(unique(cohort_ev[, c("id", "time", "evid")])))

# rxSolve returns OBSERVATION rows only and drops `evid` from its output, so
# there is nothing to filter here -- but assert the row count rather than
# assuming it, so a future change to build_events() fails loudly.
cohort_sim <- rxode2::rxSolve(mod, cohort_ev, keep = c("CRCL", "ALB"),
                              returnType = "data.frame") |>
  mutate(
    day        = (time - 23) / 24 + 1,
    sim_censor = pmax(sim, LLOQ / 2)   # LLOQ/2 substitution; see note above
  )

stopifnot(
  nrow(cohort_sim) == N_SUBJ * length(TDM_TIMES),
  !anyNA(cohort_sim$Cc), !anyNA(cohort_sim$sim_censor)
)
```

```{r cohort-vpc}
vpc <- cohort_sim |>
  group_by(day) |>
  summarise(
    Q10 = quantile(sim_censor, 0.10),
    Q50 = quantile(sim_censor, 0.50),
    Q90 = quantile(sim_censor, 0.90),
    .groups = "drop"
  )

ggplot(vpc, aes(day, Q50)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(
    x = "Maintenance day", y = "23 h trough (mg/L)",
    title = "Simulated trough distribution, 200 virtual HSCT subjects",
    caption = paste(
      "Median with 10th-90th percentile band, including between-subject and",
      "residual variability, censored at LLOQ/2. Comparable in construction to",
      "Chae 2026 Figure 3A (pcVPC within 720 h)."
    )
  )
```

```{r cohort-assertions}
# Assertions on a simulated cohort: CENTRE and ROBUST QUANTILES only. The
# typical-value day-28 trough from the deterministic solve above is the anchor;
# the cohort median must sit near it, and the spread must be wide but finite.
typ_d28    <- tail(trough_by_day$trough, 1)
cohort_d28 <- cohort_sim |> filter(day == max(day))

med_ratio <- median(cohort_d28$sim_censor) / typ_d28
p10_p90   <- quantile(cohort_d28$sim_censor, c(0.10, 0.90))

stopifnot(
  # The median of a cohort with large IIV is not the typical value, but with
  # log-normal etas on CL and V2 it must stay within a factor of ~2 of it.
  med_ratio > 0.4, med_ratio < 2.0,
  # A 35% CV on CL plus a 50% proportional residual gives a wide but bounded
  # 10-90 spread; a mis-transcribed omega scale (e.g. reading 0.118 as a CV)
  # collapses or explodes this ratio.
  p10_p90[[2]] / p10_p90[[1]] > 2.5,
  p10_p90[[2]] / p10_p90[[1]] < 20
)

tibble::tibble(
  Quantity = c("Typical-value day-28 trough (mg/L)",
               "Cohort median day-28 trough (mg/L)",
               "Cohort 10th percentile (mg/L)",
               "Cohort 90th percentile (mg/L)"),
  Value = signif(c(typ_d28, median(cohort_d28$sim_censor), p10_p90[[1]], p10_p90[[2]]), 3)
) |>
  knitr::kable(caption = "Day-28 trough: typical value versus the simulated cohort.")
```

# Assumptions and deviations

* **Residual-error scale.** Chae 2026 Table 4 prints `sigma_add` = 0.672 and
  `sigma_prop` = 0.245 without stating a scale. They are encoded here as NONMEM
  `$SIGMA` **variances**, so the packaged SDs are `sqrt(0.672)` = 0.8198 mg/L
  and `sqrt(0.245)` = 0.4950 (a 49.5% proportional residual). Three things
  support that reading, all shown in *Reading the random-effect scale* above:
  Table 4's omega rows are demonstrably variances; Table 3 flags the one place
  the authors transformed anything ("as CV%") and leaves the residual header
  bare; and the printed RSE reproduces the printed bootstrap CI on the variance
  scale. **The competing reading** -- that 0.245 is already an SD, i.e. a 24.5%
  proportional residual -- is the more usual published magnitude and matches the
  24.3% of the sibling Korean teicoplanin model `Wi_2017_teicoplanin`. A user
  who prefers it can set `propSd = 0.245` and `addSd = 0.672`. Only the width of
  the simulated observation distribution changes; every typical-value result in
  this vignette is unaffected.
* **Infusion duration is not reported.** A 0.5 h infusion is assumed. The
  sensitivity check above shows 23 h troughs move by under 2% across durations
  from a near-bolus to 2 h, so nothing TDM-relevant depends on the choice.
* **Maintenance-dose timing.** Chae 2026 says three 400 mg loading doses at 12 h
  intervals then 400 mg once daily, without giving the interval between the last
  loading dose and the first maintenance dose. Doses are placed here at 0, 12,
  24 h and then q24h from 48 h, i.e. the first maintenance dose falls one full
  day after the last loading dose. The paper also notes the actual dose was at
  the physician's discretion and the standard regimen was not always applied.
* **Covariate distributions.** eGFR and albumin are drawn independently from
  normal distributions matched to the Chae 2026 Table 1 means and SDs, truncated
  to positive values. The paper reports no correlation between them and no
  distributional shape beyond mean +/- SD; in reality the two are likely
  mildly correlated through nutritional and inflammatory status, so the
  simulated cohort's clearance spread is a slight over-estimate.
* **Covariates are time-fixed.** Chae 2026 captures eGFR and albumin as
  per-patient demographic variables and reports no within-subject time course,
  so both are held constant per subject. In routine HSCT care both move during
  a teicoplanin course.
* **Structural misspecification is inherited, not introduced.** The authors
  report that a three-compartment model fitted the pooled data better on OFV,
  residual normality and homoscedasticity, but destabilised the SSE loop; they
  selected two compartments for estimation stability and explicitly acknowledge
  the resulting misspecification. The packaged model reproduces their published
  two-compartment choice.
* **The equation normalisers differ slightly from the Table 1 means.** The
  clearance equation normalises to eGFR 113.31 and albumin 3.10 g/dL, while
  Table 1 prints cohort means of 113.38 and 3.04. The equation's constants are
  used because `theta5` and `theta6` were estimated against them.
* **Scope.** Patients on renal replacement therapy or massive fluid therapy were
  excluded from the analysis, so the model has no dialysis-clearance term and
  should not be applied to those patients. The paper also reports that
  predictive performance degrades beyond 720 h from the first dose, where the
  TDM data become sparse; simulations in this vignette stay inside that window.
* **No non-paper-derived parameter values.** Every `ini()` entry comes from
  Chae 2026 Table 4, with Table 3 and the printed clearance equation as
  corroborating sources. No value was digitised from a figure, supplied by
  correspondence, or carried from an upstream model.
