---
title: "Canakinumab (Chakraborty 2012)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Canakinumab (Chakraborty 2012)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

```{r setup}
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(PKNCA)
```

# Canakinumab population PK / IL-1b binding model

Canakinumab is a fully human anti-interleukin-1b (IL-1b) IgG1/k monoclonal
antibody developed by Novartis (commercial name: Ilaris). Its primary
regulatory indication is cryopyrin-associated periodic syndromes (CAPS), a
group of inherited NALP3-inflammasome-driven autoinflammatory disorders.
Chakraborty et al. (2012) integrated data from six clinical studies (n = 233
subjects) into a single population pharmacokinetic-binding model with
quasi-steady-state (QSS) drug-target binding, structurally adapted from the
Hayashi 2007 omalizumab framework. Two physical compartments (central and
peripheral) each carry three species: free drug, free IL-1b, and the
canakinumab-IL-1b complex. Drug, ligand, and complex share volumes; complex
clearance is set equal to free-drug clearance.

* Citation: `r rxode2::rxode(readModelDb("Chakraborty_2012_canakinumab"))$reference`
* Article: <https://doi.org/10.2165/11599820-000000000-00000> (open access)

# Population

The pooled model-building dataset contains 233 subjects across six clinical
studies (Chakraborty 2012 Table I and Table III):

| Study              | Population              | Approx. n contribution | Dosing                                             |
|--------------------|-------------------------|------------------------|----------------------------------------------------|
| CACZ885B2101       | Healthy + mild asthma   | 37                     | 1, 3, 10 mg/kg IV; multiple-dose                   |
| CACZ885A1101       | Japanese healthy        | 30                     | 1, 3 mg/kg IV; 600 mg IV; 150, 300 mg SC; combo    |
| CACZ885A2101       | Rheumatoid arthritis    | 37                     | 0.3, 1, 3, 10 mg/kg IV; multiple-dose              |
| CAZC885A2202       | Plaque psoriasis        | 10                     | 150 mg SC single                                   |
| CACZ885A2102       | CAPS (phase II)         | 34                     | 10 mg/kg IV; 150 mg / 2 mg/kg SC                   |
| CACZ885D2304       | CAPS (phase III)        | 85                     | 150 mg SC q8w (registration regimen)               |

Median typical CAPS patient: 70 kg, 43 g/L serum albumin, 34 years. Sex
distribution across the analysis cohort was 135 male / 98 female (42% female;
Chakraborty 2012 Section 5.3 page e15). Age range 4-74 years, including
paediatric CAPS subjects. The same metadata is available via
`readModelDb("Chakraborty_2012_canakinumab")$population`.

This vignette focuses on the typical adult CAPS patient because CAPS is the
regulatory indication driving the registration dosing regimen
(150 mg SC q8w). Per-population point estimates for Caucasian healthy
volunteers, Japanese healthy volunteers, RA, mild asthma, and psoriasis
cohorts are reported in Chakraborty 2012 Table IV; they are tabulated in the
"Per-population estimates" section below.

# Source trace

Per-parameter origin is recorded as inline comments next to each `ini()`
entry in `inst/modeldb/specificDrugs/Chakraborty_2012_canakinumab.R`. The
table below collects them in one place for review.

| Equation / parameter        | Value (paper)                | Value (file)             | Source                          |
|-----------------------------|------------------------------|--------------------------|---------------------------------|
| `lcl` (CLD, CAPS, 70 kg)    | 0.174 L/d                    | log(0.174)               | Table IV                        |
| `lvc` (VC at 70 kg)         | 3.30 L                       | log(3.30)                | Table IV                        |
| `lvp` (VP at 70 kg)         | 2.71 L                       | log(2.71)                | Table IV                        |
| `lpsd` (PSD)                | 0.429 L/d                    | log(0.429)               | Table IV                        |
| `lka` (Sp2/0, 34 y)         | 0.269 1/d                    | log(0.269)               | Table IV                        |
| `lcll` (CLL, CAPS)          | 14.2 L/d                     | log(14.2)                | Table IV                        |
| `lrli` (RLI, CAPS)          | 9.57 ng/d                    | log(9.57)                | Table IV                        |
| `lkd` (KD, CAPS)            | 1.07 nmol/L                  | log(1.07)                | Table IV                        |
| `lpsl` (PSL)                | 0.386 L/d                    | log(0.386)               | Table IV                        |
| `lfdepot` (F1, Sp2/0)       | 70.0% (logit-back-calc)      | logit(0.70) = 0.847      | Table IV (footnote a)           |
| `e_wt_cl`                   | 0.695                        | 0.695                    | Table IV (Covariates)           |
| `e_alb_cl`                  | -0.916                       | -0.916                   | Table IV; equation 11           |
| `e_wt_vc`                   | 0.684                        | 0.684                    | Table IV; equation 10           |
| `e_wt_vp`                   | 0.798                        | 0.798                    | Table IV; equation 10           |
| `e_age_ka`                  | -0.555                       | -0.555                   | Table IV                        |
| `etalcl`                    | omega^2 = 0.0859 (CV ~29%)   | 0.0859                   | Table IV (CLD IIV)              |
| `etalvc`                    | omega^2 = 0.0589 (CV ~24%)   | 0.0589                   | Table IV (VC IIV)               |
| `etalvp`                    | omega^2 = 0.0817 (CV ~29%)   | 0.0817                   | Table IV (VP IIV)               |
| `etalpsd`                   | omega^2 = 0.280 (CV ~53%)    | 0.280                    | Table IV (PSD IIV)              |
| `etalka`                    | omega^2 = 0.406 (CV ~64%)    | 0.406                    | Table IV (ka IIV; on NS0 row)   |
| `etalcll`                   | omega^2 = 0.371 (CV ~61%)    | 0.371                    | Table IV (CLL IIV)              |
| `etalrli`                   | omega^2 = 0.261 (CV ~51%)    | 0.261                    | Table IV (RLI IIV)              |
| `etalkd`                    | omega^2 = 0.395 (CV ~63%)    | 0.395                    | Table IV (KD IIV)               |
| `etalpsl`                   | omega^2 = 0.254 (CV ~50%)    | 0.254                    | Table IV (PSL IIV)              |
| `propSd` (canakinumab)      | sigma^2 = 0.0527 (CV ~23%)   | sqrt(0.0527) = 0.230     | Table IV; eq 8 (Y = log(IPRED) + e) |
| `propSd_totalIL1b`          | sigma^2 = 0.0840 (CV ~29%)   | sqrt(0.0840) = 0.290     | Table IV; eq 8                  |
| Eq. 1 dA/dt = -ka * A       | n/a                          | n/a                      | Methods page e7                 |
| Eq. 2 dTDC/dt drug central  | n/a                          | n/a                      | Methods page e7-e8              |
| Eq. 3 dTDP/dt drug periph.  | n/a                          | n/a                      | Methods page e8                 |
| Eq. 4 dTLC/dt ligand cent.  | n/a                          | n/a                      | Methods page e8                 |
| Eq. 5 dTLP/dt ligand per.   | n/a                          | n/a                      | Methods page e8                 |
| Eq. 7 QSS quadratic for XC  | n/a                          | n/a                      | Methods page e8                 |
| MW canakinumab ~150 kDa     | 150 kDa                      | 149.117 kDa              | Section 4.2.2 page e10          |
| MW IL-1b 17000 ng/nmol      | 17 kDa                       | 17.000 kDa               | Eq. 5 footnote                  |
| Centred power covariate eq. | y = y70 * (WT/70)^exp        | y * (WT/70)^exp          | Equation 10 page e15            |
| Albumin scaling             | CLD = CLD * (alb/43)^-0.92   | encoded in `cl`          | Equation 11 page e15            |

The IIV CV percentages reported by Chakraborty 2012 are well approximated by
sqrt(omega^2) (e.g. variance 0.0859 -> 29% CV, variance 0.406 -> 64% CV), so
the ini() variance entries can be read directly from Table IV without a
log-normal CV-to-variance conversion. The residual error is reported as
"additive on log-scale" (eq. 8: Y = log(IPRED) + e); for small sigma this is
indistinguishable from a proportional model with propSd = sigma, the
encoding used here.

# Per-population point estimates from Table IV

The published model treated study population (CAPS, Caucasian HV, Japanese
HV, RA, asthma, psoriasis) as a categorical covariate on CLD, CLL, RLI, and
KD (Section 4.2.2 page e9). The model in this package uses CAPS as the
typical population (regulatory indication and dosing-driver). For
completeness, the per-population point estimates are tabulated below; they
are not encoded as covariates in `inst/modeldb/specificDrugs/Chakraborty_2012_canakinumab.R`.

| Population         | CLD (L/d) | CLL (L/d) | RLI (ng/d) | KD (nmol/L) |
|--------------------|-----------|-----------|------------|-------------|
| CAPS (typical)     | 0.174     | 14.2      | 9.57       | 1.07        |
| Caucasian HV       | 0.137     | 13.5      | 4.50       | 1.01        |
| Japanese HV        | 0.169     | 31.0      | 4.72       | 0.354       |
| Rheumatoid arth.   | 0.195     | 22.9      | 8.06       | 0.227       |
| Mild asthma        | 0.168     | 24.9      | 6.18       | 0.543       |
| Psoriasis          | 0.185     | 19.0      | 7.19       | 0.360       |

Source: Chakraborty 2012 Table IV. To simulate any population other than
CAPS, override the corresponding `lcl`, `lcll`, `lrli`, `lkd` values in the
returned `rxode2` model.

# Virtual cohort

Original observed data are not publicly available. The cohort below
reproduces the registered CAPS regimen of 150 mg SC every 8 weeks plus the
single-dose IV reference arm at 10 mg/kg used for the Table III NCA
comparison. Body weight, age, and serum albumin are sampled to approximate
the typical adult CAPS patient.

```{r cohort}
set.seed(20250509)

n_per_arm <- 50L
obs_times <- c(0.0833, 0.5, 1, 2, 3, 5, 7, 10, 14, 21, 28, 42, 56, 70, 84)

build_sc_arm <- function(n, dose_mg, regimen, id_offset) {
  ev <- rxode2::et(amt = dose_mg, cmt = "depot", id = id_offset + seq_len(n)) |>
    rxode2::et(obs_times, cmt = "Cc")
  ev <- as.data.frame(ev)
  ev$regimen <- regimen
  ev$WT  <- exp(rnorm(n, log(70), 0.15))[match(ev$id, unique(ev$id))]
  ev$AGE <- runif(n, 18, 65)             [match(ev$id, unique(ev$id))]
  ev$ALB <- pmax(30, rnorm(n, 43, 4))    [match(ev$id, unique(ev$id))]
  ev
}

build_iv_arm <- function(n, dose_mg, regimen, id_offset) {
  # central is in nmol; convert mg dose to nmol via 1000/MWX (149.117 kDa).
  amt_nmol <- dose_mg / 149.117 * 1000
  ev <- rxode2::et(amt = amt_nmol, cmt = "central", dur = 2 / 24,
                   id = id_offset + seq_len(n)) |>
    rxode2::et(obs_times, cmt = "Cc")
  ev <- as.data.frame(ev)
  ev$regimen <- regimen
  ev$WT  <- exp(rnorm(n, log(70), 0.15))[match(ev$id, unique(ev$id))]
  ev$AGE <- runif(n, 18, 65)             [match(ev$id, unique(ev$id))]
  ev$ALB <- pmax(30, rnorm(n, 43, 4))    [match(ev$id, unique(ev$id))]
  ev
}

events <- dplyr::bind_rows(
  build_sc_arm(n_per_arm, dose_mg = 150, regimen = "150 mg SC SD",
               id_offset =   0L),
  build_iv_arm(n_per_arm, dose_mg = 700, regimen = "10 mg/kg IV SD",
               id_offset = 200L)
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod <- readModelDb("Chakraborty_2012_canakinumab")
sim <- rxode2::rxSolve(mod, events = events, keep = c("regimen", "WT", "AGE", "ALB"))
sim_df <- as.data.frame(sim)
```

# Replicate Figure 1 - 150 mg SC single-dose canakinumab profile (CAPS)

```{r figure-1, fig.cap = "Replicates the canakinumab profile of Figure 1 of Chakraborty 2012 (typical CAPS adult, 150 mg SC single dose)."}
sim_typical <- rxode2::rxSolve(rxode2::zeroRe(mod),
                               events = dplyr::filter(events, regimen == "150 mg SC SD",
                                                      id == 1L))

ggplot(as.data.frame(sim_typical), aes(time, Cc)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time (days)",
       y = "Canakinumab (ug/mL)",
       title = "Typical CAPS patient: 150 mg SC single dose",
       caption = "Replicates the 150 mg SC profile of Chakraborty 2012 Figure 1.") +
  theme_minimal()
```

```{r figure-1-il1b, fig.cap = "Total IL-1b time course in the same simulation; reflects accumulation of canakinumab-bound complex following SC dosing (Chakraborty 2012 Figure 1, lower panel)."}
ggplot(as.data.frame(sim_typical), aes(time, totalIL1b)) +
  geom_line(linewidth = 0.7, colour = "darkred") +
  labs(x = "Time (days)",
       y = "Total IL-1b (pg/mL)",
       title = "Typical CAPS patient: total IL-1b after 150 mg SC dose",
       caption = "Total IL-1b accumulates because complex CL = free-drug CL << free-IL-1b CL.") +
  theme_minimal()
```

# PKNCA validation

```{r pknca}
sim_nca <- sim_df |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, regimen)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, regimen) |>
  dplyr::mutate(
    amt   = ifelse(regimen == "10 mg/kg IV SD", 700, amt),
    route = ifelse(regimen == "10 mg/kg IV SD", "intravascular", "extravascular")
  )

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | regimen + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | regimen + id,
                             route = "route")

intervals <- data.frame(
  start      = 0,
  end        = 84,
  cmax       = TRUE,
  tmax       = TRUE,
  auclast    = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- suppressMessages(PKNCA::pk.nca(nca_data))
nca_summary <- summary(nca_res)
knitr::kable(nca_summary,
             caption = "Simulated NCA parameters by dosing regimen (PKNCA, n = 50 per arm).")
```

## Comparison against Chakraborty 2012 Table III

| Regimen          | Parameter        | Paper (mean +/- SD)             | Simulation (median, this vignette) |
|------------------|------------------|---------------------------------|------------------------------------|
| 150 mg SC SD     | Cmax (ug/mL)     | 15.9 +/- 3.52 (adult CAPS)      | see PKNCA table above              |
| 150 mg SC SD     | Tmax (d)         | 6.98 (1.92-14.0)                | see PKNCA table above              |
| 150 mg SC SD     | t1/2 (d)         | 26.1 +/- 7.31                   | see PKNCA table above              |
| 150 mg SC SD     | CL/F (L/d)       | 0.228 +/- 0.0597                | implicit in dose / AUCinf          |
| 10 mg/kg IV SD   | Cmax (ug/mL)     | 149 +/- 45.4 (n=4 adult CAPS)   | see PKNCA table above              |
| 10 mg/kg IV SD   | t1/2 (d)         | 31.2 +/- 3.39                   | see PKNCA table above              |

Differences within ~20% of the paper's mean values are expected because the
simulation uses CAPS-typical covariate values whereas Table III aggregates
over patients of varying weight. Larger discrepancies (e.g. > 30% on Cmax,
half-life shorter than 20 days, AUC differing by an order of magnitude)
should trigger a re-audit of the source-trace table - do not tune the
parameters in `inst/modeldb/`.

# Steady-state and accumulation

The paper notes (page e11) that the registration regimen of 150 mg SC q8w
reaches steady state at approximately 130 days (~5 half-lives) with about
1.3-fold accumulation in a typical 70 kg CAPS adult. The chunk below
verifies the typical-value model behaviour against this benchmark.

```{r ss-accumulation}
ev_q8w <- rxode2::et(amt = 150, cmt = "depot", ii = 56, until = 280) |>
  rxode2::et(seq(0, 280, by = 7), cmt = "Cc")
ev_q8w$WT  <- 70
ev_q8w$AGE <- 34
ev_q8w$ALB <- 43

sim_ss <- as.data.frame(rxode2::rxSolve(rxode2::zeroRe(mod), ev_q8w))

cmax_first <- max(sim_ss$Cc[sim_ss$time <= 56])
cmax_ss    <- max(sim_ss$Cc[sim_ss$time >= 224 & sim_ss$time <= 280])
cat(sprintf("Cmax cycle 1: %.2f ug/mL\n", cmax_first))
cat(sprintf("Cmax cycle 5: %.2f ug/mL  (accumulation factor %.2f vs paper expected ~1.3)\n",
            cmax_ss, cmax_ss / cmax_first))
```

```{r figure-q8w, fig.cap = "Typical-value canakinumab profile under the registered 150 mg SC q8w regimen reaching steady state by ~5 half-lives (~130 days) per Chakraborty 2012 page e11."}
ggplot(sim_ss, aes(time, Cc)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time (days)", y = "Canakinumab (ug/mL)",
       title = "150 mg SC q8w typical CAPS profile",
       caption = "Accumulation factor reported in paper: ~1.3 (page e11).") +
  theme_minimal()
```

# Assumptions and deviations

- **Population covariate not encoded.** Chakraborty 2012 Table IV reports
  per-population CLD, CLL, RLI, and KD point estimates for CAPS, Caucasian
  HV, Japanese HV, RA, mild asthma, and psoriasis cohorts. This package
  encodes the typical CAPS values; per-population values are tabulated
  above for downstream users who wish to override. Encoding population
  type as a covariate column would have required new canonical entries
  (DIS_RA, DIS_CAPS, DIS_ASTHMA) in `inst/references/covariate-columns.md`
  and was deferred. CAPS is the regulatory indication and the population
  whose dosing regimen drove the model's clinical use.

- **Cell-line covariate not encoded.** The paper estimates separate ka
  and F1 for the NS0 (early development) and Sp2/0 (commercial) cell
  lines; this file uses the Sp2/0 values (ka = 0.269 1/d, F1 = 0.70)
  matching the commercial Ilaris product. The NS0 alternatives are
  ka = 0.299 1/d and F1 = 0.633 (Chakraborty 2012 Table IV) for users
  reproducing pre-2010 study data.

- **Bioavailability F1 encoded as a fixed parameter on the logit scale.**
  The paper coded F1 via a logit transformation to constrain it to (0, 1)
  during estimation (page e9); no IIV was reported on F1, and the
  back-calculated point estimate (70.0% +/- 8.25%) was reported on the
  natural scale via simulation (Table IV footnote a). The model file
  uses `lfdepot <- log(0.70 / 0.30)` and `f(depot) <- 1 / (1 + exp(-lfdepot))`
  to preserve this structural choice while allowing an end user to
  perturb F1 explicitly if needed.

- **Initial conditions for endogenous IL-1b set analytically.** The paper
  describes the dosing as starting from steady state without specifying
  the t = 0 amounts of IL-1b in the central and peripheral compartments.
  This package sets `central_il1b(0) = RLI * VC / (MWE * 1000 * CLL)`
  and `peripheral1_il1b(0) = (VP/VC) * central_il1b(0) * (1 + CLL/PSL)`,
  which is the analytical solution of the no-drug steady state
  (eq. 4 with TDC = TDP = XC = XP = 0).

- **Residual error encoded as proportional.** Chakraborty 2012 used
  Y = log(IPRED) + e (eq. 8) with sigma^2 reported on the log scale.
  For small sigma this is well approximated by `Y ~ prop(propSd)` with
  propSd = sigma; the largest sigma reported is 0.290 (IL-1b) which is
  within the regime where the prop approximation is faithful to ~1%.

- **Inter-parameter IIV correlations.** Not reported in the paper. All
  IIVs are encoded as independent / diagonal here.

- **MW(canakinumab) = 149.117 kDa** is taken from the standard human
  IgG1/k mass for canakinumab; the paper rounds to "~150 kDa" in
  Section 4.2.2.

- **Cohort covariate distributions.** WT was sampled lognormally around
  70 kg (CV 15%); AGE uniform 18-65; ALB normal 43 g/L SD 4 g/L. None
  of these distributions were reported in the paper; they bracket the
  paper's typical-CAPS reference values.
