---
title: "Efavirenz (Chala 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Efavirenz (Chala 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Chala_2023_efavirenz"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1002/psp4.12951>
* Supplement (Appendix S1, the NONMEM control stream; Tables S1-S3; Figures S1-S3):
  <https://www.ebi.ac.uk/europepmc/webservices/rest/PMC10272302/supplementaryFiles>

Chala 2023 developed a one-compartment population pharmacokinetic model with
first-order absorption for oral efavirenz in antiretroviral-therapy-naive
HIV-1-infected Ethiopian children. Apparent oral clearance CL/F carries five
multiplicative factors -- allometric body weight, `CYP2B6*6` genotype,
`ABCB1` c.4036A>G (rs3842) genotype, a genotype-gated autoinduction step, and a
latent two-class mixture -- while apparent volume V/F scales linearly with
weight. Absorption rate `ka` and the interindividual variability on V/F and `ka`
could not be identified from the sparse design and were fixed.

# Population

```{r population, echo = FALSE}
# `readModelDb()` returns the model *function*, so the population metadata is
# reached through the parsed ui built above, not by calling the function.
pop <- ui$population

# The paragraphs below are hand-written, so assert the numbers they quote
# against the model's own metadata; the two cannot drift apart silently.
stopifnot(
  identical(pop$species, "human"),
  pop$n_subjects == 100,
  pop$n_studies == 1,
  pop$sex_female_pct == 42
)
```

One hundred combination-antiretroviral-therapy-naive children aged 3-16 years
were enrolled from seven hospital ART centres in the Oromia and Southern
Nations, Nationalities and Peoples regional states of Ethiopia (Chala 2023
Table 1). Median age was 9 years (IQR 6-13), median weight 22.05 kg
(IQR 16.8-28.25) and 42% were female. Baseline liver and renal function were
broadly normal (median eGFR 95.5 mL/min/1.73m^2, median albumin 3.8 mg/dL) and
CD4 counts were immunocompetent (median 330 cells/dL); 15% had active pulmonary
tuberculosis and 69% were receiving cotrimoxazole prophylaxis.

Genotype frequencies (Table 1) are the ones that make this cohort informative:
`CYP2B6*6` was `*1/*1` in 45%, `*1/*6` in 45% and `*6/*6` in 8%, and
`ABCB1` c.4036A>G (rs3842) was G/G in 17% versus G/A or A/A in 80%.

554 efavirenz plasma concentrations were collected over one year in a two-arm
design: 13 children gave rich samples at 0, 2.5, 16 and 24 h after the *first*
dose (9 repeated at week 8), and 87 children gave a single mid-dose sample
8-16 h post dose at weeks 4, 8, 12, 24 and 48. That sparseness is the reason
`ka` and two of the three IIV terms are fixed rather than estimated.

The same information is available programmatically via
`rxode2::rxode(readModelDb("Chala_2023_efavirenz"))$population`.

# Source trace

Every `ini()` entry carries an in-file comment naming its source location in
`inst/modeldb/specificDrugs/Chala_2023_efavirenz.R`. They are collected here for
review. Point estimates are taken from supplement **Table S3 run32** -- the final
run, the only one carrying all twelve thetas -- which prints the same values as
main-text Table 2 to more significant figures.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` (CL/F at the 22 kg reference) | 4.30 L/h | Table 2 `CL (L/h)` 4.30 (RSE 13%, bootstrap 3.20-5.30); Equation 1 `4.3(CLpop)` |
| `lvc` (V/F at the 22 kg reference) | 123.80 L | Table 2 `V c (L)` 123.80 (RSE 10%, bootstrap 103.30-163.80) |
| `lka` | 0.776 1/h, fixed | Table 2 `K a (/h) 0.78` + footnote "Fixed to this value"; unrounded value in Results para. 7 and abstract; Appendix S1 `$THETA (0.776) FIX ; KA` |
| `e_wt_cl` | 0.75, fixed | Results para. 3 step 3 `CLi = CLpop * (WT/22)^0.75`; Discussion para. 3 "fixed to theoretical values"; Appendix S1 `(WT/22)**0.75` |
| `e_wt_vc` | 1, fixed | Results para. 3 step 3 `Vi = Vpop * (WT/22)`; Appendix S1 `TVV = THETA(2)*(WT/22)` |
| `e_cyp2b6_6het_cl` | 0.7245 | Table 2 "Fraction of typical CL ... CYP2B6*1/*6" 0.72; Table S3 run32 `CP2B6S1S6` |
| `e_cyp2b6_6hom_cl` | 0.2823 | Table 2 "Fraction of typical CL ... CYP2B6*6/*6" 0.28; Table S3 run32 `CP2B6S6S6` |
| `e_snp_abcb1_rs3842_a_cl` | 1.452 | Table 2 "Fold of typical CL ... ABCB1.rs3842 G/A or A/A" 1.45; Table S3 run32 `ABCB1RS3842` |
| `e_autoind_cyp2b6_6wt_cl` | 0.1235 | Table 2 "Proportional increase in CL from >= 12-weeks ... CYP2B6*1/*1" 0.12; Table S3 run32 `CP2B6S1S1WEEK12` |
| `e_autoind_cyp2b6_6het_cl` | 0.2298 | Table 2 "Proportional increase in CL from >= 8weeks ... CYP2B6*1/*6" 0.23; Table S3 run32 `CP2B6S1S6WEEK8` |
| `e_mix_slow_elim_efv_cl` | 0.2829 | Table 2 "Fraction of typical of CL for the subpopulation" 0.28; Table S3 run32 `MIXPOP` |
| mixture class probability | 0.07511 | Table 2 "Proportion of an unknown subpopulation" 0.075; Appendix S1 `$MIX ... P(1) = THETA(11)`. Carried in `covariateData$MIX_SLOW_ELIM_EFV$notes`, not in `ini()` |
| `etalcl` | 0.118127 | Table 2 "Interindividual variability for CL (%CV)" 35.4; Table S3 run32 `IIVCL` 0.3541. `omega^2 = log(CV^2 + 1)` per Methods para. 2 (`CV% = 100 * sqrt(exp(omega^2) - 1)`). Scale confirmed by the same row's bootstrap 0.105 (95% CI 0.05-0.169), which is on the variance scale and brackets 0.118127 |
| `propSd` | 0.4969 | Table 2 "Proportional residual error (%CV)" 50%; Table S3 run32 `PROP` |
| `addSd` | 0.00028 ug/mL, fixed | Appendix S1 `$THETA (0.00028) FIX ; ADD`. Not reported in main-text Table 2 -- see Errata |
| CL/F equation (product of indicator-powered factors) | n/a | Equation 1; Appendix S1 `TVCL = THETA(1) * CP2B6CL * ABCB1RS3842CL * (WT/22)**0.75 * CLWKCP2B61 * CLWKCP2B62 * CLMIX` |
| `d/dt(depot)`, `d/dt(central)` | n/a | Results para. 2 "one-compartment model parameterized in oral clearance, oral volume of distribution and absorption rate constant"; Appendix S1 `$SUBROUTINE ADVAN2 TRANS=2` |
| Residual error form | n/a | Appendix S1 `$ERROR: W = SQRT(ADD**2 + PROP**2*IPRED**2); Y = IPRED + W*ERR(1)` with `$SIGMA 1 FIX` |

# Structural verification: reproducing Equation 1

Chala 2023 Equation 1 writes individual clearance as a product of factors, each
raised to a 0/1 indicator power:

```
CLi = ( 4.3 * (WT/22)^0.75 * 0.72^f1 * 0.28^f2 * 1.45^f3
            * 1.12^f4 * 1.23^f5 * 0.28^f6 ) * exp(eta)
```

The checks below are **deterministic** -- they compare typical values from the
packaged model against closed-form arithmetic, so exact tolerances are correct
here and are deliberately tight.

```{r structural-scenarios}
mod <- readModelDb("Chala_2023_efavirenz")
mod_typical <- rxode2::zeroRe(mod)

# One scenario per row. `week` is weeks on efavirenz-based ART; the canonical
# T_FIRSTDOSE column carries it in hours (168 h per week).
scen <- tibble::tribble(
  ~scenario,                              ~WT,  ~cyp,  ~aCar, ~week, ~mix,
  "Reference: 22 kg, *1/*1, G/G, week 1",  22,     0L,     0L,     1,   0L,
  "CYP2B6 *1/*6",                          22,     1L,     0L,     1,   0L,
  "CYP2B6 *6/*6",                          22,     2L,     0L,     1,   0L,
  "ABCB1 rs3842 G/A or A/A",               22,     0L,     1L,     1,   0L,
  "*1/*1 at week 8 (no step yet)",         22,     0L,     0L,     8,   0L,
  "*1/*1 at week 12 (step applies)",       22,     0L,     0L,    12,   0L,
  "*1/*6 at week 4 (no step yet)",         22,     1L,     0L,     4,   0L,
  "*1/*6 at week 8 (step applies)",        22,     1L,     0L,     8,   0L,
  "*6/*6 at week 48 (never steps)",        22,     2L,     0L,    48,   0L,
  "Latent slow-eliminator subpopulation",  22,     0L,     0L,     1,   1L,
  "70 kg adult-size extrapolation",        70,     0L,     0L,     1,   0L,
  "10 kg child",                           10,     0L,     0L,     1,   0L
)

# Closed-form Equation 1 (typical value; exp(eta) = 1 under zeroRe).
eq1_cl <- function(WT, cyp, aCar, week, mix) {
  4.30 * (WT / 22)^0.75 *
    0.7245^(cyp == 1L) *
    0.2823^(cyp == 2L) *
    1.452^aCar *
    (1 + 0.1235)^(cyp == 0L & week >= 12) *
    (1 + 0.2298)^(cyp == 1L & week >= 8) *
    0.2829^mix
}
eq1_vc <- function(WT) 123.80 * (WT / 22)

# Build a one-dose / one-observation event table per scenario and solve.
ev_struct <- scen |>
  mutate(id = dplyr::row_number()) |>
  tidyr::crossing(tibble(time = c(0, 12), evid = c(1L, 0L),
                         amt = c(300, NA_real_), cmt = c("depot", "central"))) |>
  mutate(
    SNP_CYP2B6_RS3745274_T_COUNT = cyp,
    SNP_ABCB1_RS3842_A_CARRIER   = aCar,
    MIX_SLOW_ELIM_EFV            = mix,
    T_FIRSTDOSE                  = week * 168
  ) |>
  arrange(id, time, desc(evid)) |>
  as.data.frame()

sim_struct <- rxode2::rxSolve(mod_typical, ev_struct,
                              keep = c("scenario", "WT", "cyp", "aCar",
                                       "week", "mix")) |>
  as.data.frame()

struct_tab <- sim_struct |>
  group_by(scenario, WT, cyp, aCar, week, mix) |>
  summarise(cl_model = mean(cl), vc_model = mean(vc), .groups = "drop") |>
  mutate(
    cl_closed = eq1_cl(WT, cyp, aCar, week, mix),
    vc_closed = eq1_vc(WT),
    cl_pct    = 100 * (cl_model - cl_closed) / cl_closed,
    vc_pct    = 100 * (vc_model - vc_closed) / vc_closed
  ) |>
  arrange(match(scenario, scen$scenario))

# Deterministic identity: the packaged model must reproduce Equation 1 to
# machine precision. This is NOT a cohort statistic, so an exact bound is the
# right gate (see known-vignette-failure-patterns.md pattern 12, which applies
# to simulated cohorts and explicitly not to closed-form identities).
stopifnot(
  max(abs(struct_tab$cl_pct)) < 1e-8,
  max(abs(struct_tab$vc_pct)) < 1e-8
)

struct_tab |>
  select(scenario, cl_model, cl_closed, cl_pct, vc_model, vc_closed) |>
  dplyr::rename(
    "Scenario"              = scenario,
    "CL/F model (L/h)"      = cl_model,
    "CL/F Equation 1 (L/h)" = cl_closed,
    "Difference (%)"        = cl_pct,
    "V/F model (L)"         = vc_model,
    "V/F closed form (L)"   = vc_closed
  ) |>
  knitr::kable(
    digits  = c(0, 4, 4, 10, 3, 3),
    caption = "Typical CL/F and V/F from the packaged model against Chala 2023 Equation 1 evaluated by hand."
  )
```

Two rows above are independent checks rather than restatements of the same
arithmetic:

* The **70 kg extrapolation** is the one number Chala 2023 reports on a scale
  other than its own 22 kg reference. Discussion paragraph 5 states the typical
  value as `10.24 (7.6-12.6) L/h/70 kg`, contrasting it with Bienczak et al.'s
  21.6 L/h/70 kg. That figure is not used anywhere in building the model file,
  so reproducing it tests the clearance estimate, the reference weight and the
  allometric exponent jointly.
* The `*6/*6` row tests the abstract's claim that clearance is "reduced by
  ... 72%" in `CYP2B6*6/*6`, and the `*1/*6` row the companion "reduced by 28%".

```{r published-cl-70kg}
cl70 <- struct_tab$cl_model[struct_tab$scenario == "70 kg adult-size extrapolation"]
pub70 <- 10.24  # Chala 2023 Discussion para. 5: "typical value (95% CI) = 10.24 [7.6-12.6] L/h/70 kg"

reduction <- struct_tab |>
  filter(scenario %in% c("CYP2B6 *1/*6", "CYP2B6 *6/*6")) |>
  mutate(pct_reduction = 100 * (1 - cl_model / 4.30))

claims <- tibble::tibble(
  Claim = c(
    "CL/F at 70 kg = 10.24 L/h (Discussion para. 5)",
    "CL/F reduced by 28% in CYP2B6*1/*6 (abstract)",
    "CL/F reduced by 72% in CYP2B6*6/*6 (abstract)"
  ),
  Units     = c("L/h", "% reduction", "% reduction"),
  Published = c(10.24, 28, 72),
  Model     = c(cl70, reduction$pct_reduction[1], reduction$pct_reduction[2])
) |>
  mutate(
    `Difference (relative %)` = 100 * (Model - Published) / Published,
    `Difference (absolute)`   = Model - Published
  )

# These are deterministic (zeroRe) closed-form values, so the only residual is
# the paper's own rounding -- but the two kinds of claim need different gates.
#
# The 70 kg clearance is printed to four significant figures, so a relative
# bound is right: 0.5% admits the rounding while a wrong allometric exponent or
# reference weight (which move it by tens of percent) still fails.
#
# The two "reduced by N%" claims come from fractions printed to two decimal
# places (0.72 and 0.28) and then rounded to a whole percent, so the honest gate
# is an ABSOLUTE bound in percentage points, not a relative one: the unrounded
# 0.7245 gives a 27.55% reduction against a published "28%", which is 0.45
# percentage points of rounding but 1.6% in relative terms. One percentage point
# covers the rounding; a mis-transcribed genotype factor moves these by tens of
# percentage points and still fails.
stopifnot(
  abs(claims$`Difference (relative %)`[1]) < 0.5,
  max(abs(claims$`Difference (absolute)`[2:3])) < 1
)

knitr::kable(claims, digits = 3,
             caption = "Model against the three CL/F statements Chala 2023 makes outside its own parameter table.")
```

## Genotype-gated autoinduction

Efavirenz induces its own metabolism. Chala 2023 Results paragraph 5 reports
that the induction becomes statistically significant at a *different* treatment
week in each `CYP2B6*6` genotype: from week 12 in `*1/*1`, from week 8 in
`*1/*6`, and not at all in `*6/*6`.

```{r figure-autoinduction}
weeks <- c(0, 1, 4, 8, 12, 24, 48)
auto <- tidyr::crossing(week = weeks, cyp = 0:2) |>
  mutate(
    Genotype = factor(cyp, 0:2,
                      c("CYP2B6 *1/*1", "CYP2B6 *1/*6", "CYP2B6 *6/*6")),
    cl = eq1_cl(22, cyp, 0L, week, 0L)
  ) |>
  group_by(Genotype) |>
  mutate(`CL/F relative to week 1` = cl / cl[week == 1]) |>
  ungroup()

ggplot(auto, aes(week, `CL/F relative to week 1`, colour = Genotype)) +
  geom_step(direction = "hv", linewidth = 0.9) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = weeks) +
  labs(x = "Weeks on efavirenz-based ART", y = "CL/F relative to week 1",
       title = "Genotype-gated autoinduction of efavirenz CL/F",
       caption = "Reproduces Chala 2023 Results paragraph 5 / Table 2 rows 9-10.") +
  theme_bw()

# Deterministic step sizes straight from Table 2.
step <- auto |>
  select(Genotype, week, `CL/F relative to week 1`) |>
  tidyr::pivot_wider(names_from = week, values_from = `CL/F relative to week 1`,
                     names_prefix = "wk")
stopifnot(
  isTRUE(all.equal(step$wk8[1],  1.0000, tolerance = 1e-10)),  # *1/*1 no step at week 8
  isTRUE(all.equal(step$wk12[1], 1.1235, tolerance = 1e-10)),  # *1/*1 steps at week 12
  isTRUE(all.equal(step$wk4[2],  1.0000, tolerance = 1e-10)),  # *1/*6 no step at week 4
  isTRUE(all.equal(step$wk8[2],  1.2298, tolerance = 1e-10)),  # *1/*6 steps at week 8
  isTRUE(all.equal(step$wk48[3], 1.0000, tolerance = 1e-10))   # *6/*6 never steps
)
```

# Virtual cohort

Original observed data are not publicly available. The cohort below approximates
the Monte-Carlo design Chala 2023 used for its Figure 3 (Methods, final
paragraph): virtual children stratified by weight band, with `CYP2B6*6` genotype
held fixed within each stratum, `ABCB1` rs3842 drawn at 80% G/A-or-A/A and the
latent slow-eliminator class drawn at the estimated 7.5% probability.

Doses are the **modal dose actually received in each weight band**, from
supplement Table S1. Only the four bands whose modal dose agrees with the
SUSTIVA label are simulated; the 25-32.5 kg and 32.5-40 kg bands are omitted
because their modal received dose (600 mg) is above the label recommendation --
Chala 2023 notes that "a few individuals received higher doses than
recommended", and including those bands would inflate the exposure claims below
rather than test them.

```{r cohort}
# set.seed() fixes the covariate draws below (they are drawn in R). It does NOT
# fix rxode2's eta draws, whose streams are partitioned per solver thread, so a
# CI runner draws different etas than this machine. Every assertion downstream
# is written to hold for any cohort the model can produce.
set.seed(20230612)

n_per_arm <- 100L  # 12 arms; well under the 200-per-arm cap

bands <- tibble::tribble(
  ~band,        ~wt_lo, ~wt_hi, ~dose_mg,
  "7.5-15 kg",     7.5,   15.0,   200,   # Table S1: 200 mg in 66.7% of the band
  "15-20 kg",     15.0,   20.0,   250,   # Table S1: 250 mg in 55.6%
  "20-25 kg",     20.0,   25.0,   300,   # Table S1: 300 mg in 61.1%
  ">40 kg",       40.0,   60.0,   600    # Table S1: 600 mg in 100%
) |>
  mutate(band = factor(band, levels = band))

genos <- tibble::tibble(
  cyp      = 0:2,
  Genotype = factor(c("CYP2B6 *1/*1", "CYP2B6 *1/*6", "CYP2B6 *6/*6"),
                    levels = c("CYP2B6 *1/*1", "CYP2B6 *1/*6", "CYP2B6 *6/*6"))
)

# Steady state at week 24 of therapy: 14 once-daily doses (t = 0, 24, ..., 312),
# then observations across the final interval. Efavirenz t1/2 here is about
# 20 h, so 13 days of dosing is well over 15 half-lives.
n_dose  <- 14L
tau     <- 24
t_last  <- (n_dose - 1L) * tau          # 312 h
wk_at_start <- 24                        # weeks on ART at the start of the window

subjects <- tidyr::crossing(bands, genos) |>
  mutate(arm = paste(band, Genotype, sep = " | ")) |>
  tidyr::uncount(n_per_arm) |>
  mutate(
    id   = dplyr::row_number(),
    WT   = runif(dplyr::n(), wt_lo, wt_hi),
    SNP_CYP2B6_RS3745274_T_COUNT = cyp,
    SNP_ABCB1_RS3842_A_CARRIER   = rbinom(dplyr::n(), 1L, 0.80),
    MIX_SLOW_ELIM_EFV            = rbinom(dplyr::n(), 1L, 0.075)
  )

dose_rows <- subjects |>
  tidyr::crossing(time = seq(0, t_last, by = tau)) |>
  mutate(evid = 1L, amt = dose_mg, cmt = "depot")

obs_rows <- subjects |>
  tidyr::crossing(time = seq(t_last, t_last + tau, by = 1)) |>
  mutate(evid = 0L, amt = NA_real_, cmt = "central")

events <- bind_rows(dose_rows, obs_rows) |>
  # T_FIRSTDOSE is treatment duration, not time after dose: it keeps rising
  # across the record and does not reset at each dosing event.
  mutate(T_FIRSTDOSE = wk_at_start * 168 + time) |>
  arrange(id, time, desc(evid)) |>
  as.data.frame()

# No id/time/evid record is repeated. Note the deliberate absence of `unique()`
# here: de-duplicating first would make the check vacuously true. The 312 h row
# appears twice by design (the last dose and the first observation of the
# window), which the `evid` column keeps distinct.
stopifnot(!anyDuplicated(events[, c("id", "time", "evid")]))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod, events = events,
  keep = c("arm", "band", "Genotype", "WT", "dose_mg",
           "SNP_ABCB1_RS3842_A_CARRIER", "MIX_SLOW_ELIM_EFV")
) |>
  as.data.frame()

stopifnot(nrow(sim) > 0, !all(is.na(sim$Cc)))
```

# Replicate Figure 3

Chala 2023 Figure 3 reports population summary statistics of the efavirenz
concentration 12 h after dose, by dosing weight band and `CYP2B6*6` genotype.
Results paragraph 8 states three findings, all reproduced below:

1. "EFV concentrations at 12-h after dose are comparable across the dosing
   weight bands."
2. "subjects with `CYP2B6*6/*6` have relatively higher EFV concentrations
   ... with greater than 80% of those subjects having EFV concentrations
   greater than 4 ug/mL."
3. "greater than 80% of subjects with `CYP2B6*1/*1` or `CYP2B6*1/*6` who
   receive EFV dosing according to the SUSTIVA label, are predicted to have EFV
   concentration greater than or equal to 1 ug/mL."

```{r figure-3}
c12 <- sim |>
  filter(time == t_last + 12) |>
  select(id, arm, band, Genotype, WT, dose_mg, C12 = Cc)

stopifnot(nrow(c12) == nrow(subjects), !anyNA(c12$C12))

ggplot(c12, aes(band, C12, fill = Genotype)) +
  geom_boxplot(outlier.size = 0.5, position = position_dodge(width = 0.8)) +
  geom_hline(yintercept = c(1, 4), linetype = "dashed", colour = "grey30") +
  scale_y_log10() +
  labs(x = "Dosing weight band", y = "Efavirenz concentration 12 h post dose (ug/mL)",
       title = "Steady-state 12-h efavirenz concentration by weight band and CYP2B6*6 genotype",
       caption = paste("Replicates Figure 3 of Chala 2023. Dashed lines mark the 1 and",
                       "4 ug/mL thresholds discussed in the paper.")) +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r figure-3-claims}
band_medians <- c12 |>
  group_by(Genotype, band) |>
  summarise(median_C12 = median(C12), .groups = "drop_last") |>
  summarise(band_spread = max(median_C12) / min(median_C12), .groups = "drop")

pct_over_4 <- c12 |>
  filter(Genotype == "CYP2B6 *6/*6") |>
  summarise(pct = 100 * mean(C12 > 4)) |>
  pull(pct)

pct_over_1 <- c12 |>
  filter(Genotype != "CYP2B6 *6/*6") |>
  summarise(pct = 100 * mean(C12 >= 1)) |>
  pull(pct)

fig3 <- tibble::tibble(
  Claim = c(
    "12-h concentrations comparable across weight bands (max/min of band medians, within genotype)",
    "% of CYP2B6*6/*6 subjects above 4 ug/mL",
    "% of CYP2B6*1/*1 or *1/*6 subjects at or above 1 ug/mL"
  ),
  `Chala 2023` = c("comparable", "> 80%", "> 80%"),
  Model = c(
    sprintf("%.2f-%.2f fold", min(band_medians$band_spread), max(band_medians$band_spread)),
    sprintf("%.1f%%", pct_over_4),
    sprintf("%.1f%%", pct_over_1)
  )
)

knitr::kable(fig3, caption = "Chala 2023 Figure 3 / Results paragraph 8 claims against the packaged model.")

# Bounds chosen with headroom over what a cohort draw can move (pattern 12 of
# known-vignette-failure-patterns.md). The band-spread bound of 1.5 is a
# magnitude claim about "comparable", not a race between two noisy statistics;
# a mis-transcribed allometric exponent takes the 7.5-15 kg / >40 kg ratio well
# past 2. The two percentage bounds sit 5-10 points below the paper's own ">80%"
# so a different eta draw cannot flip them, while a wrong genotype fold-factor
# (which moves these by tens of points) still fails them.
stopifnot(
  max(band_medians$band_spread) < 1.5,
  pct_over_4 > 70,
  pct_over_1 > 75
)
```

The `CYP2B6*6/*6` stratum sits about three-and-a-half-fold above the other two
genotypes, which is the paper's central clinical message: on label dosing these
children are systematically over-exposed relative to the 1-4 ug/mL therapeutic
range, and the Discussion proposes an approximately three-fold dose reduction
for them.

Both threshold claims are stated in the paper as lower bounds ("greater than
80%"), and the packaged model clears them with room to spare -- around 97% on
each. Part of that margin is a design difference rather than a model
disagreement: the four weight bands simulated here are the label-consistent
ones, whereas Chala 2023's Figure 3 also spans the smallest weight bands and
draws its virtual weights from NHANES rather than uniformly. The direction and
the ordering across genotypes are what the replication establishes; the exact
percentage is not a published number to match.

```{r figure-profile}
sim |>
  mutate(tad = time - t_last) |>
  group_by(tad, Genotype) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = median(Cc),
            Q95 = quantile(Cc, 0.95), .groups = "drop") |>
  ggplot(aes(tad, Q50, colour = Genotype, fill = Genotype)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.18, colour = NA) +
  geom_line(linewidth = 0.9) +
  scale_y_log10() +
  labs(x = "Time after the last dose (h)", y = "Efavirenz concentration (ug/mL)",
       title = "Steady-state efavirenz profile by CYP2B6*6 genotype (all weight bands pooled)",
       caption = "Median with 5th-95th percentile band; week 24 of therapy.") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# PKNCA validation

Chala 2023 reports **no** non-compartmental parameters -- neither Cmax, Tmax,
AUC nor half-life appears in the paper or the supplement -- so there is nothing
to compare a simulated NCA table against. The NCA below is therefore run against
**exact internal identities** on a typical-value (no-IIV, no-residual-error)
profile, which is a stricter gate than a 20%-tolerance comparison would be:

* at steady state, `AUC(0-tau) = Dose / (CL/F)` exactly for a linear model;
* the terminal half-life must equal `log(2) * (V/F) / (CL/F)`, because
  `kel` (0.0347 1/h at the 22 kg reference) is far below `ka = 0.776 1/h` and no
  flip-flop occurs.

Note that the half-life is not constant across weight bands: `V/F` scales with
`WT^1` and `CL/F` with `WT^0.75`, so `log(2) * V/F / (CL/F)` grows as `WT^0.25`
-- from about 15 h in the smallest band to about 22 h in the largest.

```{r pknca}
nca_subj <- bands |>
  mutate(
    id  = dplyr::row_number(),
    WT  = (wt_lo + wt_hi) / 2,
    arm = paste0(band, " (", dose_mg, " mg)"),
    SNP_CYP2B6_RS3745274_T_COUNT = 0L,   # CYP2B6 *1/*1
    SNP_ABCB1_RS3842_A_CARRIER   = 0L,   # rs3842 G/G
    MIX_SLOW_ELIM_EFV            = 0L    # main population
  )

nca_events <- bind_rows(
  nca_subj |>
    tidyr::crossing(time = seq(0, t_last, by = tau)) |>
    mutate(evid = 1L, amt = dose_mg, cmt = "depot"),
  nca_subj |>
    tidyr::crossing(time = seq(t_last, t_last + tau, by = 0.25)) |>
    mutate(evid = 0L, amt = NA_real_, cmt = "central")
) |>
  mutate(T_FIRSTDOSE = wk_at_start * 168 + time) |>
  arrange(id, time, desc(evid)) |>
  as.data.frame()

sim_nca_raw <- rxode2::rxSolve(mod_typical, nca_events,
                               keep = c("arm", "dose_mg", "WT")) |>
  as.data.frame()

# PKNCA input filter: `!is.na(Cc)` ONLY. Adding `time > 0` or `Cc > 0` would
# drop the row that anchors the AUC interval and trigger the
# "Requesting an AUC range starting before the first measurement" warning.
sim_nca <- sim_nca_raw |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, arm)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id,
                             concu = "ug/mL", timeu = "h")

dose_df <- nca_events |>
  filter(evid == 1L) |>
  select(id, time, amt, arm)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id, doseu = "mg")

intervals <- data.frame(
  start     = t_last,
  end       = t_last + tau,
  cmax      = TRUE,
  tmax      = TRUE,
  cmin      = TRUE,
  auclast   = TRUE,
  cav       = TRUE,
  half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  select(arm, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

expected <- sim_nca_raw |>
  group_by(arm, dose_mg, WT) |>
  summarise(cl = mean(cl), vc = mean(vc), .groups = "drop") |>
  mutate(
    auc_expected = dose_mg / cl,             # Dose / (CL/F) at steady state
    thalf_expected = log(2) * vc / cl     # log(2) / kel
  )

nca_chk <- nca_wide |>
  left_join(expected, by = "arm") |>
  mutate(
    auc_pct   = 100 * (auclast - auc_expected) / auc_expected,
    thalf_pct = 100 * (half.life - thalf_expected) / thalf_expected
  )

nca_chk |>
  select(arm, cmax, tmax, cmin, cav, auclast, auc_expected, auc_pct,
         half.life, thalf_expected, thalf_pct) |>
  dplyr::rename(
    "Weight band (dose)"        = arm,
    "Cmax,ss (ug/mL)"           = cmax,
    "Tmax (h)"                  = tmax,
    "Cmin,ss (ug/mL)"           = cmin,
    "Cavg,ss (ug/mL)"           = cav,
    "AUC0-tau PKNCA (ug*h/mL)"  = auclast,
    "AUC0-tau = Dose/CL"        = auc_expected,
    "AUC difference (%)"        = auc_pct,
    "t1/2 PKNCA (h)"            = half.life,
    "t1/2 = log(2)*V/CL (h)"    = thalf_expected,
    "t1/2 difference (%)"       = thalf_pct
  ) |>
  knitr::kable(
    digits  = 3,
    caption = paste("Steady-state NCA on a typical-value profile against exact",
                    "closed-form identities. Chala 2023 publishes no NCA table,",
                    "so these identities are the validation target.")
  )

# Deterministic profile (zeroRe, no residual error), so the only error source is
# trapezoidal integration on a 0.25 h grid and lambda-z regression. Both are
# numerical, not stochastic, and identical on any machine -- an exact bound is
# correct here. Realised: AUC within 0.02% (all four bands -0.016 to -0.019%),
# t1/2 within 0.75% (all four bands +0.68 to +0.72%, the small positive bias
# being lambda-z fitted over a window where the absorption term has not fully
# decayed).
stopifnot(
  max(abs(nca_chk$auc_pct))   < 0.5,
  max(abs(nca_chk$thalf_pct)) < 2,
  # Efavirenz is absorption-limited nowhere near flip-flop: Tmax must land well
  # inside the interval, and Cmin,ss at the END of the interval for a lag-free
  # oral model.
  all(nca_chk$tmax > 0), all(nca_chk$tmax < 12)
)
```

Cavg,ss is `AUC0-tau / tau`, i.e. `Dose / (CL/F * 24)`. It lands at 2.5-2.9
ug/mL in every label-consistent weight band for a `CYP2B6*1/*1` child, squarely
inside the 1-4 ug/mL therapeutic range Chala 2023 cites (Discussion paragraph
4), and the narrow spread across bands is the same "comparable across weight
bands" result the Figure 3 replication shows. The 15-22 h half-life is
consistent with once-daily dosing and with the roughly 2-fold peak-to-trough
ratio in the table.

# Assumptions and deviations

## Errata and reporting conflicts in the source

* **The legend defining `f4` and `f5` under Equation 1 mislabels one genotype.**
  It reads "`f4` = if greater than 12 weeks and `CYP2B6*1/*6`; `f5` = if greater
  than or equal to 8 weeks and `CYP2B6*1/*6`", assigning `*1/*6` to *both*
  indicators. Four other places in the paper say `f4` belongs to `CYP2B6*1/*1`:
  Table 2's row label ("Proportional increase in CL from >= 12-weeks for
  subjects with `CYP2B6*1/*1`"), Results paragraph 5 ("For `CYP2B6*1/*1`
  genotype ... the difference between week 1 and week 12 onward was
  statistically significant"), Results paragraph 7 ("greater than 12 or 8 weeks
  on treatment for `CYP2B6*1/*1` or `CYP2B6*1/*6`, respectively"), and the
  abstract ("clearance was higher from weeks 8 and 12 in `CYP2B6*1/*6` and
  `CYP2B6*1/*1` genotypes, respectively"). The supplement's NONMEM control
  stream settles it outright:
  `CLWKCP2B61 = ((1+THETA(9))**WK12 * (1+THETA(9))**WK24)**CP2B6S1S1`, i.e.
  THETA(9) = 0.1235 is gated on the `*1/*1` indicator. The model file follows
  the control stream; the legend is a typographical error.

* **The autoinduction step is coded on a discrete week grid in the control
  stream and as a `>=` threshold everywhere else.** Appendix S1 uses
  `WK8 = (WEEK.EQ.8)`, `WK12 = (WEEK.EQ.12)` and `WK24 = (WEEK.GE.24)`, which is
  exactly equivalent to a `>=` threshold on the sampling grid the study actually
  used (weeks 0/1, 4, 8, 12, 24, 48) but leaves weeks 9-11 and 13-23 undefined.
  Table 2 ("from >= 12-weeks", "from >= 8 weeks"), Equation 1 and the abstract
  all describe it as a threshold, and the threshold form is the only one that is
  well defined at an arbitrary simulation time. The model file uses `>=`.

* **The additive residual-error term is absent from the main text.** Table 2
  reports only the 50% proportional term, but Appendix S1's `$ERROR` block is
  `W = SQRT(ADD**2 + PROP**2*IPRED**2)` with `$THETA (0.00028) FIX ; ADD`. The
  model file carries `addSd <- fixed(0.00028)` for fidelity to the published
  control stream. At 0.28 ng/mL it is about 56-fold below the assay LLOQ of
  15.78 ng/mL and four orders of magnitude below therapeutic concentrations, so
  it acts as a numerical stabiliser and changes no result in this vignette.

* **CL/F is printed as 4.30 in Table 2 and 4.29 in Table S3 run32.** Table 2 is
  the designated final-model table and agrees with Equation 1's `4.3(CLpop)`,
  the abstract's "4.3 L/h", the bootstrap median (4.30) and the Discussion's
  70 kg restatement, so 4.30 is used. The 0.2% discrepancy is presentational.
  Every other parameter is taken from Table S3 run32, which prints the same
  values as Table 2 to more significant figures.

* **The week-12 `*1/*1` autoinduction RSE differs between the two reports.**
  Results paragraph 5 quotes RSE = 56% for that effect, Table 2 reports 71%. The
  former describes the covariate-building step, the latter the final model. Only
  the point estimate (0.1235) enters the model file, so nothing downstream
  depends on which RSE is right.

* **No erratum exists.** Crossref reports no `update-to` / `updated-by` relation
  for `doi:10.1002/psp4.12951` and no correction notice was found.

## Encoding decisions

* **IIV on V/F and `ka` is omitted rather than written as `~ fixed(0)`.** Both
  were fixed to zero in the source (Results paragraph 7; Appendix S1
  `$OMEGA 0.227 ; IIVCL / 0 FIX ; IIVV / 0.000001 FIX ; IIVKA`). Writing them as
  zero-variance etas would make OMEGA singular and break rxode2's Cholesky
  sampler, so CL/F is the model's only random effect -- which is exactly what
  the source model does.

* **The `$MIX` block becomes a covariate column, not an estimated mixture.**
  rxode2 has no `$MIXTURE` analogue, so the latent class enters as the binary
  `MIX_SLOW_ELIM_EFV` column with the estimated class probability (0.07511)
  recorded in `covariateData$MIX_SLOW_ELIM_EFV$notes`. Set it to 0 for
  typical-value work; draw `Bernoulli(0.075)` per subject for population
  simulation, which is what Chala 2023 itself did (it rounded to 8%).

* **A new canonical covariate column was needed for `ABCB1` rs3842.** The
  register's existing `SNP_ABCB1_RS3842` is a *G*-allele-carrier indicator with
  A/A as its reference (Mukonzo 2009). Chala 2023 pools the opposite way -- G/A
  or A/A versus a G/G reference -- and because the heterozygote falls in the "1"
  group under both poolings, neither column can be derived from the other. The
  extraction registers `SNP_ABCB1_RS3842_A_CARRIER` alongside it. The two papers
  agree on the biology (both associate the G allele with higher efavirenz
  exposure); they differ only in which genotypes they pool against which
  reference.

* **`MIX_SLOW_ELIM_EFV` was registered as the sibling the register pre-named.**
  The `MIX_SLOW_ELIM_NVP` entry's Notes explicitly instruct that "future
  fast/slow CYP2B6-driven elimination mixtures for other antiretrovirals (e.g.,
  efavirenz ...) should register sibling canonicals (`MIX_SLOW_ELIM_EFV`, etc.)
  rather than reuse this entry".

* **`CYP2B6*6` is encoded through the canonical 516G>T allele-count column.**
  Chala 2023 genotyped and reports "`CYP2B6*6`" throughout and identifies it with
  c.516G>T in the Discussion, so `*1/*1`, `*1/*6` and `*6/*6` map onto
  `SNP_CYP2B6_RS3745274_T_COUNT` values 0, 1 and 2. `model()` decomposes the
  count back into the paper's indicators, matching the precedent in
  `Sanchez_2011_efavirenz.R` and `Schipani_2011_nevirapine.R`.

## Simulation assumptions

* **Weight within a band is drawn uniformly.** Chala 2023 sampled its virtual
  subjects from NHANES stratified by age and weight band; NHANES is not used
  here, so weight is uniform on each band. The `>40 kg` band is capped at 60 kg,
  a bound the paper does not state.

* **Doses are the modal dose actually received per weight band (Table S1), not
  the SUSTIVA label table.** The label's dose-by-weight table is not reproduced
  in the paper or supplement, so using it would introduce values from outside
  the source. Only the four bands whose modal received dose matches the label
  are simulated; the 25-32.5 kg and 32.5-40 kg bands are excluded because their
  modal received dose (600 mg) exceeds the label.

* **Treatment duration is fixed at week 24 across the simulated window.** All
  three genotypes' autoinduction steps have fully applied by then, so Figure 3
  is reproduced on the induced steady state. `T_FIRSTDOSE` still advances with
  simulation time within the window, as the canonical column requires.

* **Genotype is held fixed within each Figure-3 stratum**, matching the paper's
  by-genotype boxplots, rather than drawn at the cohort frequencies. `ABCB1`
  rs3842 and the mixture class are drawn at the paper's simulation
  probabilities (80% A-carrier, 7.5% slow class).

* **No parameter value in this extraction came from anywhere other than Chala
  2023's main text, Table 2, Equation 1, or supplement Appendix S1 / Tables
  S1-S3.** Nothing was digitised from a figure, obtained by correspondence, or
  carried from an upstream model.
