---
title: "Dolutegravir pediatric weight-band dosing (Chandasana 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dolutegravir pediatric weight-band dosing (Chandasana 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Chandasana_2023_dolutegravir"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1007/s40262-023-01289-5>
* Supplement (Electronic Supplementary Material, includes the final NONMEM
  control stream as Supplementary Text 1):
  <https://static-content.springer.com/esm/art%3A10.1007%2Fs40262-023-01289-5/MediaObjects/40262_2023_1289_MOESM1_ESM.docx>

Chandasana 2023 pooled the IMPAACT P1093 and ODYSSEY (PENTA 20) pediatric
dolutegravir trials into a single population PK model, and used it to confirm
the WHO weight-band doses for infants, children and adolescents down to 4 weeks
of age and 3 kg. This vignette reproduces the simulated steady-state exposures
of the paper's Table 3 and checks the model against the paper's own printed
derived quantities.

# Population

The analysis pooled 239 participants -- 151 from IMPAACT P1093 (a phase I/II
multicentre open-label single-arm trial in children and adolescents aged 4 weeks
to <18 years) and 88 from ODYSSEY (a phase II/III multicentre open-label
randomised non-inferiority trial in children aged at least 28 days and
<18 years). 240 participants provided 2714 plasma concentrations, of which 64
were excluded (21 below the limit of quantification, 7 outliers or very low
concentrations, 4 sample mix-ups, 13 haemolysed samples, 6 for non-adherence and
13 from one participant with multiple co-morbidities), leaving 2650 samples
(1909 intensive and 741 sparse) from 239 participants (Chandasana 2023
Sect. 3.1).

Baseline age ranged from 0.170 to 17.5 years (median 6.00), baseline weight from
3.9 to 91.0 kg, the cohort was 49.8% female, and 79.5% were Black -- the ODYSSEY
cohort being 100% Black (Chandasana 2023 Table 1). Dolutegravir was given once
daily under fasting conditions as film-coated tablets, dispersible tablets or
granules for oral suspension across the WHO weight bands.

The same information is available programmatically via
`readModelDb("Chandasana_2023_dolutegravir")()$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Chandasana_2023_dolutegravir.R`. The table
below collects them in one place for review. "Control stream" refers to the
final NONMEM model code reproduced verbatim as Supplementary Text 1 of the
Electronic Supplementary Material.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL/F at 70 kg, FMAT = 1) | 1.03 L/h (95% CI 0.980-1.07) | Table 2 |
| `lvc` (V/F at 70 kg) | 13.6 L (95% CI 13.0-14.3) | Table 2 |
| `lka` (Ka, film-coated tablet) | 0.854 1/h (95% CI 0.686-1.06) | Table 2 |
| `e_form_dt_ka` (Ka ratio, DT / granules) | 2.04 (95% CI 1.41-2.67); product 0.854 x 2.04 = 1.74 1/h | Table 2; covariate-relationship block; Table S1 |
| `e_wt_cl` | 0.455 (95% CI 0.418-0.492) | Table 2 |
| `e_wt_vc` | 0.556 (95% CI 0.514-0.598) | Table 2 |
| `tmat50` | 52.2 postmenstrual weeks, FIX | Table 2 (footnote a; Anderson & Holford 2009); control stream `THETA(8)` |
| `hill_mat` | 3.43, FIX | Table 2 (footnote a); control stream `THETA(7)` |
| `e_food_fdepot` (without regard to food) | 1.10 (95% CI 1.03-1.17) | Table 2; control stream `THETA(4)` |
| `e_form_dt_fdepot` (DT / granules) | 1.53 (95% CI 1.43-1.63) | Table 2; control stream `THETA(9)` |
| IIV block `etalcl`, `etalvc`, `etalka` | var 0.0863 / 0.0698 / 0.762; cov 0.0499, 0.0953, 0.138 | Table 2; control stream `$OMEGA BLOCK(3)` |
| `etaiov_lcl_1..4` | 0.115 each (CV 33.9%) | Table 2 (four rows); control stream `$OMEGA BLOCK(1)` + `SAME(3)` |
| `etaiov_lka_1..2` | 0.610 each (CV 91.7%) | Table 2 (two rows); control stream `$OMEGA BLOCK(1)` + `SAME(1)` |
| `propSd` / `addSd` (P1093) | var 0.0818 / 0.00164, i.e. CV 28.6% and SD 0.0405 ug/mL | Table 2 |
| `propSd_odyssey` / `addSd_odyssey` | var 0.0123 / 0.0900, i.e. CV 11.1% and SD 0.300 ug/mL | Table 2 |
| `FMAT = PMA^Hill / (PMA^Hill + TM50^Hill)`, `PMA = PNA*52 + 40` | n/a | Covariate-relationship block under Table 2; control stream `$PK` |
| `CL/F = 1.03 * (WT/70)^0.455 * FMAT`, `V/F = 13.6 * (WT/70)^0.556` | n/a | Covariate-relationship block under Table 2 |
| `F1 = 1 * 1.10^SFLAG * 1.53^FFLAG` | n/a | Control stream `$PK` (with its own comment block enumerating the four combinations); Table S1 |
| One-compartment first-order absorption and elimination | n/a | Sect. 3.2; control stream `$SUBROUTINES ADVAN2 TRANS2` |
| Combined proportional + additive error, per study | n/a | Control stream `$ERROR` |

## The two effects that are ratios, not absolute values

Two rows of Table 2 are easy to misread, and both are settled unambiguously by
the control stream and by Table S1:

* **`KA~DT and granules = 2.04` is a multiplier, not a Ka.** The control stream
  forms `KAFORM = THETA(17)**FORMK` and `KA = EXP(MU_3 + ETA(3)) * IOVKA *
  KAFORM`. The dispersible-tablet Ka is therefore `0.854 x 2.04 = 1.74 1/h`,
  which the paper spells out in the covariate-relationship block and tabulates
  in Table S1 as 1.74 (95% CI 1.20-2.28).
* **The 1.10 food effect is not film-coated-tablet-specific.** The control
  stream forms `F1 = 1 * THETA(4)**SFLAG * THETA(9)**FFLAG`, and its own
  comments enumerate fasted FCT = 1, fasted DT = `THETA(9)`, fed FCT =
  `THETA(4)`, fed DT = `THETA(9) * THETA(4)`. Table S1 accordingly gives
  F for the dispersible tablet without regard to food as 1.68 = 1.10 x 1.53.
  Table 2 lists a fed estimate on the FCT row only because that row is the
  reference level.

Both are checked numerically below.

## Errata noted while transcribing

* **Sect. 3.2 prints the weight exponent on V/F as 0.566**, while Table 2 gives
  0.556 with a 95% CI of 0.514-0.598 that is centred on 0.556. Table S1 also
  gives 0.556. The tabulated value is used.
* **The covariate-relationship block prints the confidence interval for
  F without regard to food, DT/granules as "1.68 (1.47-0.91)".** The upper
  bound is a transposition: the stated arithmetic is 1.03 x 1.43 = 1.47 to
  1.17 x 1.63 = 1.91. Only the point estimate 1.68 enters the model.
* **Table 3's Cmax column is not mutually consistent with its own AUC0-24
  column** under the published model: in 8 of the 11 QD arms the reported median
  Cmax exceeds the largest value attainable at an infinite absorption rate given
  the reported median AUC0-24, and the remaining 3 sit within 3% of that
  ceiling. This is demonstrated numerically below. The AUC0-24 and C24 columns
  are consistent with each other and with the model, and are reproduced here.
* **Table 1's categorical counts do not sum to the column N** (e.g. metal-cation
  products "Absent 123 (81.5)" plus "Present 45 (29.8)" for N = 151). These are
  time-varying flags counted per record rather than mutually exclusive
  per-subject categories. No model parameter depends on them -- every covariate
  outside weight, age, formulation and food was fixed to no effect in the final
  model (`THETA(10)`-`THETA(16)` are all `0 FIX` or `1 FIX`).

# Deterministic checks against the paper's printed derived quantities

Before simulating a cohort, confirm the packaged model reproduces the four
derived quantities the paper prints for itself. These are exact algebraic
identities, so they are asserted tightly.

```{r derived-quantities}
mod <- readModelDb("Chandasana_2023_dolutegravir")
ini_tbl <- ui$iniDf
th <- function(nm) ini_tbl$est[match(nm, ini_tbl$name)]

ka_fct    <- exp(th("lka"))
ka_dt     <- exp(th("lka") + th("e_form_dt_ka"))
f_fct_fed <- exp(th("e_food_fdepot"))
f_dt_fast <- exp(th("e_form_dt_fdepot"))
f_dt_fed  <- exp(th("e_food_fdepot") + th("e_form_dt_fdepot"))

derived <- tibble::tibble(
  Quantity  = c("Ka, film-coated tablet (1/h)",
                "Ka, dispersible tablet / granules (1/h)",
                "F, film-coated tablet without regard to food",
                "F, fasted dispersible tablet / granules",
                "F, dispersible tablet / granules without regard to food"),
  Model     = c(ka_fct, ka_dt, f_fct_fed, f_dt_fast, f_dt_fed),
  Published = c(0.854, 1.74, 1.10, 1.53, 1.68),
  Source    = c("Table 2", "Covariate block / Table S1", "Table 2",
                "Table 2", "Covariate block / Table S1")
) |>
  dplyr::mutate(`% difference` = 100 * (Model - Published) / Published)

knitr::kable(derived, digits = c(0, 3, 3, 0, 2),
             caption = "Model-derived formulation and food effects vs the values Chandasana 2023 prints for itself.")

# Exact algebra against printed 3-significant-figure values: the only slack is
# the paper's own rounding, so 0.5% is generous and a mis-encoded ratio (e.g.
# 2.04 read as an absolute Ka) would show up as a 17% or 134% miss.
stopifnot(max(abs(derived$`% difference`)) < 0.5)
```

The maturation function is likewise checked against the two statements the paper
makes about it: `TM50` is 52.2 postmenstrual weeks, which the paper equates to
12 weeks of postnatal age, and `FMAT` must be at half of its maximum there.

```{r maturation}
pma_weeks <- function(pna_years) pna_years * 52 + 40
fmat <- function(pma_wk) pma_wk^th("hill_mat") / (pma_wk^th("hill_mat") + th("tmat50")^th("hill_mat"))

# 12 weeks postnatal = 12/52 years -> PMA = 52 weeks; TM50 = 52.2 weeks.
stopifnot(abs(pma_weeks(12 / 52) - 52) < 1e-9)
stopifnot(abs(fmat(th("tmat50")) - 0.5) < 1e-12)
# Maturation is essentially complete by school age (FMAT = 0.9986 at 6 years)
# and all but exact in adolescence (0.99996 at 18 years). Deterministic
# algebra, so these bounds sit just under the exact values.
stopifnot(fmat(pma_weeks(6)) > 0.998, fmat(pma_weeks(18)) > 0.9999)
# ...and materially incomplete in a young infant, which is what separates the
# 10 mg (<6 months) and 15 mg (>=6 months) doses in the 6-10 kg band.
stopifnot(fmat(pma_weeks(2 / 12)) < 0.55, fmat(pma_weeks(9 / 12)) > 0.75)
```

# Virtual cohort

Original observed data are not publicly available. Chandasana 2023 Sect. 2.4
simulated 200 participants (100 male, 100 female) for each of 11 weight
band / dose / formulation combinations, sampling age uniformly from the relevant
CDC growth chart and deriving each subject's weight from that chart. The cohort
below mirrors that design at the same 200 participants per arm: age is drawn
uniformly over the age window implied by each band, weight is taken from a
piecewise-linear interpolation of the CDC 2000 sex-averaged 50th-percentile
weight-for-age curve with lognormal scatter, and the result is clamped into the
band the arm is defined by.

```{r cohort}
# rxSetSeed fixes rxode2's RNG per solver thread, not across thread counts, so
# CI draws a different cohort than a workstation does. Every assertion below is
# written to hold for any cohort this model can produce (see the notes on each).
rxode2::rxSetSeed(20230821)
set.seed(20230821)

# CDC 2000 50th-percentile weight-for-age, sexes averaged. Reference data for
# constructing the virtual cohort only -- NOT a model parameter, and not taken
# from Chandasana 2023.
cdc_age_yr <- c(1, 2, 3, 4, 6, 9, 12, 18, 24) / 12
cdc_wt_kg  <- c(4.4, 5.5, 6.3, 7.0, 7.9, 9.0, 9.9, 11.3, 12.5)
cdc_age_yr <- c(cdc_age_yr, 3:18)
cdc_wt_kg  <- c(cdc_wt_kg, 14.3, 16.3, 18.3, 20.7, 23.2, 26.0, 29.2, 32.7,
                36.7, 41.3, 46.0, 50.6, 54.6, 58.0, 60.6, 62.6)
cdc_weight_for_age <- function(age_yr) {
  stats::approx(cdc_age_yr, cdc_wt_kg, xout = age_yr, rule = 2)$y
}

# The 11 QD weight band / dose / formulation combinations of Chandasana 2023
# Table 3. Age windows are those the paper attaches to each band (Table 3 and
# Sect. 3.4); where a band carries no explicit age, the window is the age range
# over which the CDC median weight falls inside the band.
bands <- tibble::tribble(
  ~arm,                            ~wt_lo, ~wt_hi, ~age_lo, ~age_hi, ~form, ~dose,
  "3 to <6 kg (>=1 month), DT 5",     3.0,    6.0,  1/12,     4/12,  "DT",   5,
  "6 to <10 kg (1-<6 mo), DT 10",     6.0,   10.0,  1/12,     6/12,  "DT",  10,
  "6 to <10 kg (>=6 mo), DT 15",      6.0,   10.0,  6/12,     2.0,   "DT",  15,
  "10 to <14 kg, DT 20",             10.0,   14.0,  1.0,      4.0,   "DT",  20,
  "14 to <20 kg, DT 25",             14.0,   20.0,  3.0,      6.0,   "DT",  25,
  "14 to <20 kg, FCT 40",            14.0,   20.0,  3.0,      6.0,   "FCT", 40,
  "20 to <25 kg, DT 30",             20.0,   25.0,  5.5,      8.0,   "DT",  30,
  "20 to <25 kg, FCT 50",            20.0,   25.0,  5.5,      8.0,   "FCT", 50,
  "25 to <30 kg, FCT 50",            25.0,   30.0,  7.5,      9.5,   "FCT", 50,
  "30 to <35 kg, FCT 50",            30.0,   35.0,  9.0,     11.0,   "FCT", 50,
  ">=35 kg, FCT 50",                 35.0,   90.0, 11.0,     18.0,   "FCT", 50
)

n_per_arm <- 200L
n_doses   <- 14L            # 14 QD doses, i.e. > 20 elimination half-lives
t_last    <- 24 * (n_doses - 1L)

# Observation grid over the final dosing interval: fine through absorption so
# Cmax and Tmax are resolved, coarser through the terminal phase.
obs_grid <- sort(unique(c(seq(0, 6, by = 0.1), seq(6, 24, by = 0.5))))

make_arm <- function(row, id_offset) {
  age_yr <- stats::runif(n_per_arm, row$age_lo, row$age_hi)
  wt <- cdc_weight_for_age(age_yr) * exp(stats::rnorm(n_per_arm, 0, 0.12))
  wt <- pmin(pmax(wt, row$wt_lo), row$wt_hi)
  subj <- tibble::tibble(
    id  = id_offset + seq_len(n_per_arm),
    arm = row$arm,
    WT  = wt,
    # The model's PAGE column is postmenstrual age in MONTHS; the source model
    # works in postmenstrual WEEKS with PMA = PNA(years) * 52 + 40, so convert
    # back through the same 4.348125 weeks-per-month factor model() applies.
    PAGE          = (age_yr * 52 + 40) / 4.348125,
    FORM_DTG_DT   = as.numeric(row$form == "DT"),
    FED           = 0,      # dosed fasted, as in P1093 and ODYSSEY
    STUDY_ODYSSEY = 0,
    OCC           = 1,      # single steady-state occasion
    dose_mg       = row$dose
  )
  doses <- subj |>
    dplyr::mutate(time = 0, amt = dose_mg, evid = 1L, cmt = "depot",
                  addl = n_doses - 1L, ii = 24)
  obs <- subj |>
    tidyr::crossing(tad = obs_grid) |>
    dplyr::mutate(time = t_last + tad, amt = NA_real_, evid = 0L,
                  cmt = "central", addl = 0L, ii = 0)
  dplyr::bind_rows(doses, obs) |> dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- do.call(
  dplyr::bind_rows,
  lapply(seq_len(nrow(bands)), function(i) {
    make_arm(bands[i, ], id_offset = (i - 1L) * n_per_arm)
  })
)

stopifnot(nrow(dplyr::distinct(events, id)) == n_per_arm * nrow(bands))
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

```{r cohort-summary, echo = FALSE}
events |>
  dplyr::filter(evid == 1) |>
  dplyr::group_by(arm) |>
  dplyr::summarise(N = dplyr::n(),
                   `Dose (mg)` = unique(dose_mg),
                   Formulation = ifelse(unique(FORM_DTG_DT) == 1, "DT", "FCT"),
                   `Weight (kg), median` = round(stats::median(WT), 1),
                   `Weight (kg), range` = paste0(round(min(WT), 1), "-", round(max(WT), 1)),
                   `Age (y), median` = round(stats::median(PAGE * 4.348125 - 40) / 52, 2),
                   .groups = "drop") |>
  dplyr::arrange(match(arm, bands$arm)) |>
  dplyr::rename("Weight band / dose" = arm) |>
  knitr::kable(caption = "Virtual cohort by simulated arm (200 participants each).")
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod, events = events,
  keep = c("arm", "WT", "PAGE", "dose_mg", "FORM_DTG_DT")
) |>
  as.data.frame() |>
  # rxSolve returns observation records only, so no evid filter is needed (and
  # `evid` is not among the returned columns); restrict to the final interval.
  dplyr::filter(time >= t_last) |>
  dplyr::mutate(tad = time - t_last)

stopifnot(nrow(sim) == n_per_arm * nrow(bands) * length(obs_grid))
```

## Steady-state concentration-time profiles

```{r figure-profiles, fig.height = 7}
sim |>
  dplyr::group_by(arm, tad) |>
  dplyr::summarise(Q10 = stats::quantile(Cc, 0.10),
                   Q50 = stats::quantile(Cc, 0.50),
                   Q90 = stats::quantile(Cc, 0.90),
                   .groups = "drop") |>
  dplyr::mutate(arm = factor(arm, levels = bands$arm)) |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.25) +
  geom_line() +
  geom_hline(yintercept = 0.995, colour = "red", linetype = "dashed") +
  geom_hline(yintercept = 0.697, colour = "blue", linetype = "dashed") +
  facet_wrap(~arm, ncol = 3) +
  scale_y_log10() +
  labs(x = "Time after dose at steady state (h)", y = "Dolutegravir (ug/mL)",
       title = "Simulated steady-state QD profiles by WHO weight band",
       caption = paste("Median with 10th-90th percentile ribbon.",
                       "Red dashed line: adult target geometric mean C24 of 0.995 ug/mL;",
                       "blue dashed line: lower target bound 0.697 ug/mL",
                       "(Chandasana 2023 Sect. 2.3). Companion to Figure 1."))
```

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, arm, time = tad, Cc)

# A time-zero record exists naturally (the observation grid starts at the dose
# time), but guarantee it defensively so PKNCA never anchors AUC before the
# first measurement.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

stopifnot(nrow(sim_nca) > 0, all(sim_nca$Cc >= 0))

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::transmute(id, arm, time = 0, amt = dose_mg)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)

intervals <- data.frame(start = 0, end = 24,
                        cmax = TRUE, tmax = TRUE, auclast = TRUE)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(arm, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

# C24 is read directly off the profile at exactly 24 h post-dose rather than
# from PKNCA's ctrough, which is defined on the interval end and is NA when the
# interval end does not coincide with an observation for every subject.
c24 <- sim |>
  dplyr::filter(abs(tad - 24) < 1e-8) |>
  dplyr::select(id, arm, C24 = Cc)

nca_wide <- dplyr::left_join(nca_wide, c24, by = c("id", "arm"))
stopifnot(!anyNA(nca_wide$C24), !anyNA(nca_wide$cmax), !anyNA(nca_wide$auclast))
```

## A tight internal identity: steady-state AUC must equal Dose x F / CL

With the random effects zeroed, the steady-state AUC over one dosing interval is
exactly `Dose * F / CL` for a one-compartment linear model, whatever the
absorption rate. Both sides use the same parameters, so the only difference is
numerical -- this check is therefore asserted tightly, unlike the cohort
comparisons that follow.

```{r identity-check}
# `dose` is a reserved name in the rxode2 / PKNCA pipeline, so the dose column
# is carried as `dose_mg` throughout.
typical <- bands |>
  dplyr::rename(dose_mg = dose) |>
  dplyr::mutate(id = dplyr::row_number(),
                WT = (wt_lo + wt_hi) / 2,
                age_yr = (age_lo + age_hi) / 2,
                PAGE = (age_yr * 52 + 40) / 4.348125,
                FORM_DTG_DT = as.numeric(form == "DT"),
                FED = 0, STUDY_ODYSSEY = 0, OCC = 1)

ev_typ <- dplyr::bind_rows(
  typical |> dplyr::mutate(time = 0, amt = dose_mg, evid = 1L, cmt = "depot",
                           addl = n_doses - 1L, ii = 24),
  typical |> tidyr::crossing(tad = obs_grid) |>
    dplyr::mutate(time = t_last + tad, amt = NA_real_, evid = 0L,
                  cmt = "central", addl = 0L, ii = 0)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim_typ <- rxode2::rxSolve(rxode2::zeroRe(mod), events = ev_typ,
                           keep = c("arm", "dose_mg")) |>
  as.data.frame() |>
  dplyr::filter(time >= t_last) |>
  dplyr::mutate(tad = time - t_last)

auc_identity <- sim_typ |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    # Trapezoidal AUC over the final interval on the simulation grid.
    auc_solved = sum(diff(tad) * (utils::head(Cc, -1) + utils::tail(Cc, -1)) / 2),
    # Closed form: Dose * F / CL, with cl and frel returned by the solve.
    auc_closed = unique(dose_mg) * unique(frel) / unique(cl),
    .groups = "drop"
  ) |>
  dplyr::mutate(`% difference` = 100 * (auc_solved - auc_closed) / auc_closed)

knitr::kable(auc_identity, digits = c(0, 2, 2, 3),
             caption = "Solved steady-state AUC0-24 vs the closed form Dose x F / CL (typical values, random effects zeroed).")

# Pure trapezoidal discretisation error on a 0.1 h absorption grid; both sides
# share the same drawn parameters, so this bound is deterministic and tight.
stopifnot(max(abs(auc_identity$`% difference`)) < 0.5)
```

## Comparison against the published simulated exposures

Chandasana 2023 Table 3 reports, for each of the 11 QD arms, the 10th
percentile, median and 90th percentile of simulated steady-state Cmax, C24 and
AUC0-24. The table below places the cohort simulated here beside it.

```{r published-table3}
published <- tibble::tribble(
  ~arm,                            ~stat,     ~cmax, ~C24,  ~auclast,
  "3 to <6 kg (>=1 month), DT 5",  "p10",      2.44, 0.357,  25.4,
  "3 to <6 kg (>=1 month), DT 5",  "median",   4.02, 1.07,   49.4,
  "3 to <6 kg (>=1 month), DT 5",  "p90",      6.78, 3.01,   95.4,
  "6 to <10 kg (1-<6 mo), DT 10",  "p10",      3.67, 0.392,  35.9,
  "6 to <10 kg (1-<6 mo), DT 10",  "median",   5.90, 1.24,   67.4,
  "6 to <10 kg (1-<6 mo), DT 10",  "p90",      9.57, 3.63,  127,
  "6 to <10 kg (>=6 mo), DT 15",   "p10",      4.24, 0.262,  36.4,
  "6 to <10 kg (>=6 mo), DT 15",   "median",   6.67, 0.964,  68.4,
  "6 to <10 kg (>=6 mo), DT 15",   "p90",     10.6,  3.21,  128,
  "10 to <14 kg, DT 20",           "p10",      4.30, 0.184,  34.3,
  "10 to <14 kg, DT 20",           "median",   6.61, 0.719,  63.1,
  "10 to <14 kg, DT 20",           "p90",     10.2,  2.56,  115,
  "14 to <20 kg, DT 25",           "p10",      4.63, 0.212,  37.8,
  "14 to <20 kg, DT 25",           "median",   7.17, 0.824,  69.5,
  "14 to <20 kg, DT 25",           "p90",     11.2,  2.87,  127,
  "14 to <20 kg, FCT 40",          "p10",      4.38, 0.254,  39.6,
  "14 to <20 kg, FCT 40",          "median",   6.96, 0.972,  72.6,
  "14 to <20 kg, FCT 40",          "p90",     11.0,  3.25,  132,
  "20 to <25 kg, DT 30",           "p10",      4.77, 0.227,  39.3,
  "20 to <25 kg, DT 30",           "median",   7.37, 0.881,  72.0,
  "20 to <25 kg, DT 30",           "p90",     11.4,  3.02,  133,
  "20 to <25 kg, FCT 50",          "p10",      4.73, 0.289,  43.2,
  "20 to <25 kg, FCT 50",          "median",   7.43, 1.08,   78.6,
  "20 to <25 kg, FCT 50",          "p90",     11.8,  3.62,  145,
  "25 to <30 kg, FCT 50",          "p10",      4.26, 0.272,  39.3,
  "25 to <30 kg, FCT 50",          "median",   6.74, 0.997,  71.4,
  "25 to <30 kg, FCT 50",          "p90",     10.6,  3.32,  131,
  "30 to <35 kg, FCT 50",          "p10",      3.94, 0.256,  36.5,
  "30 to <35 kg, FCT 50",          "median",   6.20, 0.944,  66.6,
  "30 to <35 kg, FCT 50",          "p90",      9.78, 3.10,  121,
  ">=35 kg, FCT 50",               "p10",      3.07, 0.233,  29.2,
  ">=35 kg, FCT 50",               "median",   4.93, 0.814,  54.0,
  ">=35 kg, FCT 50",               "p90",      7.93, 2.59,   99.1
)

simulated <- nca_wide |>
  dplyr::group_by(arm) |>
  dplyr::summarise(dplyr::across(c(cmax, C24, auclast),
                                 list(p10 = ~stats::quantile(.x, 0.10),
                                      median = ~stats::median(.x),
                                      p90 = ~stats::quantile(.x, 0.90))),
                   .groups = "drop") |>
  tidyr::pivot_longer(-arm, names_to = c("param", "stat"), names_sep = "_") |>
  tidyr::pivot_wider(names_from = param, values_from = value)

cmp <- dplyr::inner_join(simulated, published, by = c("arm", "stat"),
                         suffix = c("_sim", "_pub")) |>
  dplyr::mutate(
    `Cmax % diff`     = 100 * (cmax_sim - cmax_pub) / cmax_pub,
    `C24 % diff`      = 100 * (C24_sim - C24_pub) / C24_pub,
    `AUC0-24 % diff`  = 100 * (auclast_sim - auclast_pub) / auclast_pub
  )
stopifnot(nrow(cmp) == nrow(published))
```

```{r published-table3-render, echo = FALSE}
cmp |>
  dplyr::mutate(arm = factor(arm, levels = bands$arm),
                stat = factor(stat, levels = c("p10", "median", "p90"))) |>
  dplyr::arrange(arm, stat) |>
  dplyr::transmute(
    "Weight band / dose"       = arm,
    "Statistic"                = stat,
    "Cmax sim (ug/mL)"         = cmax_sim,
    "Cmax pub (ug/mL)"         = cmax_pub,
    "Cmax % diff"              = `Cmax % diff`,
    "C24 sim (ug/mL)"          = C24_sim,
    "C24 pub (ug/mL)"          = C24_pub,
    "C24 % diff"               = `C24 % diff`,
    "AUC0-24 sim (ug*h/mL)"    = auclast_sim,
    "AUC0-24 pub (ug*h/mL)"    = auclast_pub,
    "AUC0-24 % diff"           = `AUC0-24 % diff`
  ) |>
  knitr::kable(digits = c(0, 0, 2, 2, 1, 3, 3, 1, 1, 1, 1),
               caption = "Simulated vs published (Chandasana 2023 Table 3) steady-state exposures, 11 QD weight band / dose / formulation arms.")
```

The AUC0-24 and C24 columns reproduce; the Cmax column does not, and the next
section shows that it cannot be reproduced by **any** absorption rate constant.
Cmax is therefore recorded as a known deviation and excluded from the gate,
rather than the gate being widened until it passes.

```{r published-table3-gate}
med <- cmp |> dplyr::filter(stat == "median")

# The two cohorts differ by a PHYSICAL mechanism, not just numerical noise: the
# paper sampled age and weight from the CDC charts with its own (unpublished)
# per-band age windows, and CL/F depends on weight^0.455 and on a maturation
# term that is steep in infancy. A cohort whose median weight sits a kilogram
# off the paper's therefore shifts exposure by several percent even with an
# identical model. So gate on the CENTRE across arms and on a robust envelope,
# per the guidance in CLAUDE.md -- never on the extreme arm.
#
# Realised on this cohort: median +5.0%, 80th percentile of |diff| 7.2%, worst
# arm 20.1% (the 6-10 kg / 1-<6 month arm, where the unpublished age window
# matters most because maturation is steepest there). A mis-transcribed
# clearance, dose or unit moves every arm together by tens of percent and
# breaks these bounds immediately.
stopifnot(
  abs(stats::median(med$`AUC0-24 % diff`)) < 15,
  stats::quantile(abs(med$`AUC0-24 % diff`), 0.8) < 25
)

# C24 is the most cohort-sensitive of the three: it sits ~20 h into the terminal
# phase, so it compounds the weight- and maturation-driven CL/F difference over
# the whole interval and amplifies the AUC discrepancy roughly two- to
# three-fold (realised median +21%). The paper's dose-selection target is its
# GEOMETRIC MEAN, so gate on that rather than on a per-arm percentage.
gm <- function(x) exp(mean(log(x)))
c24_gm <- nca_wide |>
  dplyr::group_by(arm) |>
  dplyr::summarise(c24_geomean = gm(C24), .groups = "drop") |>
  dplyr::mutate(formulation = ifelse(grepl("FCT", arm), "FCT", "DT"))

knitr::kable(c24_gm, digits = 3,
             caption = "Geometric mean steady-state C24 by arm; the paper's adult-matching target is 0.995 ug/mL, its target range 0.697-2.260, and its individual lower limit 0.500.")

# Chandasana 2023 Sect. 3.4 / Abstract: "The predicted geometric mean trough
# concentration was comparable to the adult value ... for all weight bands at
# recommended doses", with per-band medians of 0.814-1.08 (FCT) and 0.719-1.24
# (DT) ug/mL against a target of 0.995. Assert the paper's claim -- every arm
# clears the individual lower limit of 0.500 ug/mL and the cohort centres on the
# adult target -- rather than an exact match to any one published number.
# Realised: arm geometric means 0.75-1.71, overall geometric mean 1.04 (+4.2%
# against the 0.995 target).
stopifnot(all(c24_gm$c24_geomean > 0.500),
          all(c24_gm$c24_geomean < 2.500),
          abs(gm(c24_gm$c24_geomean) - 0.995) / 0.995 < 0.35)
```

## Known deviation: the published Cmax column is not reachable at any absorption rate

Simulated Cmax runs uniformly about 19% below Chandasana 2023 Table 3 (-8% to
-21% across the 11 arms) while AUC0-24 agrees within a few percent. A uniform
peak deficit with a correct area means the simulated profiles are flatter than
the published ones, which is a statement about absorption, not about clearance.

That can be tested without knowing the absorption rate at all. For a
one-compartment model with first-order absorption at steady state, Cmax rises
monotonically with `ka`, and its supremum as `ka -> Inf` (an instantaneous
input, i.e. an intravenous bolus) is

```
Cmax <= (F * D / V) / (1 - exp(-kel * tau))
      = AUCtau * kel / (1 - exp(-kel * tau))
```

The right-hand side involves only the dosing-interval AUC and `kel = CL/V` --
**no absorption parameter**. So it can be evaluated using the paper's *own*
published median AUC0-24 together with the model's `kel`, and it bounds what any
version of this model could have produced.

```{r cmax-ceiling}
kel_by_id <- sim |>
  dplyr::group_by(arm, id) |>
  dplyr::summarise(kel = dplyr::first(kel), .groups = "drop")

ceiling_check <- dplyr::left_join(nca_wide, kel_by_id, by = c("id", "arm")) |>
  dplyr::mutate(ceiling = auclast * kel / (1 - exp(-24 * kel)))

# Sanity check on our own cohort: the bound is algebra, so every simulated
# subject must satisfy it. This goes red if the dosing interval, the NCA window
# or the steady-state assumption is wrong -- it is a real check, not a tautology.
stopifnot(all(ceiling_check$cmax <= ceiling_check$ceiling * (1 + 1e-6)))

published_ceiling <- ceiling_check |>
  dplyr::group_by(arm) |>
  dplyr::summarise(kel_median = stats::median(kel), .groups = "drop") |>
  dplyr::inner_join(dplyr::filter(published, stat == "median"), by = "arm") |>
  dplyr::mutate(
    `Ceiling at ka = Inf` = auclast * kel_median / (1 - exp(-24 * kel_median)),
    `Published / ceiling` = cmax / `Ceiling at ka = Inf`,
    `Reachable?` = ifelse(`Published / ceiling` > 1, "no", "only as ka -> Inf")
  )
```

```{r cmax-ceiling-render, echo = FALSE}
published_ceiling |>
  dplyr::mutate(arm = factor(arm, levels = bands$arm)) |>
  dplyr::arrange(arm) |>
  dplyr::transmute(
    "Weight band / dose"            = arm,
    "kel (1/h)"                     = kel_median,
    "Published AUC0-24 (ug*h/mL)"   = auclast,
    "Published Cmax (ug/mL)"        = cmax,
    "Ceiling at ka = Inf (ug/mL)"   = `Ceiling at ka = Inf`,
    "Published / ceiling"           = `Published / ceiling`,
    "Reachable?"                    = `Reachable?`
  ) |>
  knitr::kable(digits = c(0, 4, 1, 2, 2, 3, 0),
               caption = "Chandasana 2023 Table 3 median Cmax against the largest Cmax the published model can produce, computed from the paper's own median AUC0-24 and the model's kel. A ratio above 1 is unattainable at any absorption rate.")
```

```{r cmax-ceiling-gate}
n_impossible <- sum(published_ceiling$`Published / ceiling` > 1)
cat("Arms whose published median Cmax exceeds the ka = Inf ceiling: ",
    n_impossible, " of ", nrow(published_ceiling), "\n", sep = "")
cat("Ratio range: ", sprintf("%.3f", min(published_ceiling$`Published / ceiling`)),
    " to ", sprintf("%.3f", max(published_ceiling$`Published / ceiling`)), "\n", sep = "")

# Gate the FINDING, not a match. Every arm's published Cmax sits at or near the
# instantaneous-absorption ceiling, so the whole column requires an absorption
# rate orders of magnitude above the published 0.854 (FCT) / 1.74 (DT) 1/h.
#
# The bound is NOT cohort-independent, as an earlier comment here assumed. The
# ratios move with the eta draw, and that draw moves with the SOLVER THREAD
# COUNT, because rxSetSeed() fixes the random stream per thread rather than
# across thread counts. Measured minima: 0.9794 (1 thread), 0.9146 (2), 0.9601
# (4), 0.9739 (16) -- so a floor calibrated on any single run is fragile, and
# 0.95 duly failed at 2 threads. 0.88 sits below the whole measured spread
# while still going red on the thing this gate exists to catch: were the
# deviation resolved, an adequate absorption rate would put these ratios far
# below 1, not a few percent under it.
stopifnot(all(published_ceiling$`Published / ceiling` > 0.88))
```

Two independent reasons make this test conservative -- that is, the true
inconsistency is larger than the table shows. The paper computed its NCA on a
coarse 0, 1, 2, 3, 4, 6, 8, 24 h grid (Sect. 2.3), so its Cmax is the maximum of
eight sampled points and **understates** the true peak, while its linear-
trapezoidal AUC over that grid **overstates** the area under a convex decay.
Both biases inflate the ceiling and deflate the published Cmax, and 8 of 11 arms
still exceed the bound.

The AUC0-24 and C24 columns, by contrast, are mutually consistent and are
reproduced here, so the structural parameters and relative bioavailability are
transcribed correctly. The most likely explanation is that Table 3's Cmax column
was produced with a different absorption parameterisation from the one in
Table 2 -- the adult model of Table S1 carries Ka = 2.24 1/h against the
pediatric 0.854 -- but nothing on disk settles which, so no change is made to
the model: the packaged file follows Table 2 and the control stream, which are
the paper's definitive statements of the fitted model.

# Assumptions and deviations

* **Virtual-cohort construction.** Chandasana 2023 sampled age uniformly from
  the infantile (up to 36 months) or pediatric (24-240 months) CDC growth chart
  and derived each subject's weight from that chart, but does not publish the
  per-arm age windows it used. The windows in `bands` above are those the paper
  attaches to each arm where it states one (the 1-<6 month and >=6 month splits
  of the 6-10 kg band, and the >=1 month floor of the 3-6 kg band); for the
  remaining arms the window is the age range over which the CDC median weight
  falls inside the band. Weight is drawn from a piecewise-linear interpolation
  of the CDC 2000 sex-averaged 50th-percentile weight-for-age curve with 12%
  lognormal scatter and clamped into the band. This reference curve is
  **not** from Chandasana 2023 and is used only to build the cohort; no model
  parameter depends on it. Differences between this cohort and the paper's are
  the dominant source of the residual percentage differences in the Table 3
  comparison above.
* **Sex is not simulated.** The paper generated 100 male and 100 female
  participants per arm, but sex was not retained as a covariate in the final
  model (`THETA(13)`, `CL~Gender`, is `0 FIX` in the control stream), so it
  affects the published cohort only through the sex-specific CDC weight curves.
  A sex-averaged curve is used here instead.
* **Dosing is simulated fasted** (`FED = 0`), matching the trial protocols
  ("Dolutegravir was administered QD under fasting conditions", Sect. 2.2). The
  paper does not state the food status of its Table 3 simulations; the fasted
  reading reproduces the published AUC0-24 values, whereas the without-regard-to-
  food reading would inflate every arm by 10%.
* **A single occasion is simulated** (`OCC = 1`), so one inter-occasion draw on
  CL/F and one on Ka enter each subject, matching a single steady-state
  profile. The source model carries four CL/F occasions and two Ka occasions.
  The paper does not say whether its Table 3 simulations drew inter-occasion
  variability; including one draw is what a NONMEM `$SIMULATION` of the
  published control stream does, so that is the default here. Re-running the
  identical cohort with `OCC = 0` (inter-occasion variability off) tightens the
  agreement -- median AUC0-24 difference falls from +5.0% to +3.0% and median
  C24 from +21% to +11% -- but leaves the Cmax deficit essentially unchanged at
  -19%, which is why the Cmax deviation above is not an inter-occasion-
  variability artefact. The faithful setting is kept rather than the
  better-matching one.
* **Study-specific residual error is not exercised.** `Cc` is the individual
  prediction and carries no residual error, which matches the paper's own
  simulation of "steady-state intensive PK concentration-time profiles ... using
  individual PK parameter values" (Sect. 2.3). `STUDY_ODYSSEY = 0` is set only
  so the column exists.
* **Postmenstrual age assumes term birth.** The source model derives
  `PMA = PNA(years) * 52 + 40`, i.e. a fixed 40-week gestation. The packaged
  model takes postmenstrual age directly through the canonical `PAGE` column, so
  a user with preterm subjects can supply a true PMA; the cohort here reproduces
  the paper's term-birth assumption.
* **Granules are pooled with the dispersible tablet**, as in the source model,
  on the strength of a healthy-adult bioequivalence study (Sect. 2.3). Table 3
  simulates no granule arm.
* **The exposure-safety analysis is not reproduced.** Chandasana 2023 Sects. 2.5
  and 3.5 fit exploratory linear-regression and logistic Emax models of
  laboratory endpoints and adverse events against exposure, and report no
  correlation for any of them; no parameter estimates are published for these
  exploratory fits (Figs. 2, 3 and S3 show fitted lines and confidence bands
  only), so there is nothing to encode.
* **Alternative BID posology is not reproduced.** Table S3's BID simulations use
  the same model and parameters, differing only in the dosing regimen; they are
  reproducible from the packaged model by halving the dose and setting
  `ii = 12`.

# Relationship to the other packaged dolutegravir models

`modellib("Chandasana_2024b_dolutegravir")` is this same fit, applied without
re-estimation (NONMEM `MAXEVAL = 0`) to the IMPAACT 2019 abacavir /
dolutegravir / lamivudine fixed-dose-combination cohort and extracted from that
paper's Table 2 reproduction. Because Chandasana 2024 reproduces the formulation
effects as ratios without restating the products, prints the residual errors as
variances, and defers the random-effects covariance matrix to this paper, the
values in that file were corrected against this primary source when it was
extracted. `modellib("Zhang_2015_dolutegravir")` is the corresponding adult
treatment-naive model, which Chandasana 2023 Table S1 sets side by side with
this one.
