---
title: "Voriconazole (Chantharit 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Voriconazole (Chantharit 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Chantharit_2019_voriconazole"))$reference`
* Description: `r rxode2::rxode(readModelDb("Chantharit_2019_voriconazole"))$description`
* Article: <https://doi.org/10.1093/ofid/ofz360.1430>

```{r load-model}
mod <- readModelDb("Chantharit_2019_voriconazole")
ui <- rxode2::rxode(mod)
```

# Population

Voriconazole is a first-line agent for invasive aspergillosis, and its
pharmacokinetics are notoriously variable. Chantharit and colleagues report
what they describe as the first population pharmacokinetic analysis of
voriconazole in Thai patients: 106 patients receiving **oral** voriconazole,
with the dataset built by combining intensive pharmacokinetic sampling with
routine therapeutic-drug-monitoring trough concentrations. Estimation was by
FOCE ELS in Phoenix NLME, and the final model was checked by bootstrap
(1000 runs, 96% successful), a visual predictive check and goodness-of-fit
plots.

CYP2C19 phenotype was available for 88 of the 106 patients (43 extensive, 37
intermediate and 8 poor metabolizers) and is the paper's headline **negative**
result: it "did not influence any PK parameter during the covariate model
building", so all 106 patients were carried into the final model. What the
authors retained instead was **serum albumin** on clearance - the title's
"novel marker" - alongside gamma-glutamyltransferase on clearance and actual
body weight on volume.

The source is a one-page IDWeek 2019 poster abstract and prints **no baseline
demographics table**. Age, sex and the covariate distributions are therefore
unavailable. The only window onto the cohort's central tendency is the set of
centring constants inside the final-model equation - albumin 28 g/L, actual
body weight 55 kg, and log-gamma-glutamyltransferase 2.4 - and those are what
the virtual cohort below is built around. An albumin median of 28 g/L means the
typical patient in this cohort was hypoalbuminaemic.

The same information is available programmatically via
`readModelDb("Chantharit_2019_voriconazole")()$population`.

# Source trace

The parameter table (Table 1) and the dosing-simulation table (Table 2) are
**embedded in the abstract page as raster images** and do not appear in the
PDF text layer. They were recovered with `pdfimages -png` and read directly.
Every value below is transcribed from that image; the two structural point
estimates are independently corroborated by the Results prose.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lka` (Ka, fixed) | 1 /h | Table 1, row `Ka (fixed)`; also Results prose ("by fixing Ka value at 1.0") |
| `lcl` (thetaCL) | 7.33 L/h | Table 1, row `thetaCL (L/h)`, 95% CI 6.78-7.89; also Results prose |
| `lvc` (thetaV) | 439.69 L | Table 1, row `thetaV (L)`, 95% CI 430.62-448.77; also Results prose |
| `e_wt_vc` (theta1, fixed) | 1 | Table 1, row `theta1 (fixed)` |
| `e_alb_cl` (theta2) | 0.93 | Table 1, row `theta2`, 95% CI 0.75-1.12 |
| `e_ggt_cl` (theta3) | -0.58 | Table 1, row `theta3`, 95% CI -0.67 to -0.49 |
| `etalcl` (omega^2 CL) | 0.190 | Table 1, row `omega^2CL (omegaCL x 100)` = 0.190 (43.58) |
| `etalvc` (omega^2 V) | 0.080 | Table 1, row `omega^2V (omegaV x 100)` = 0.080 (28.28) |
| `propSd` (sigma) | 0.59 | Table 1, row `Residual variability(sigma)`, 95% CI 0.54-0.64 |
| `CL = thetaCL * (albumin/28)^theta2 * (logGGT/2.4)^theta3 * exp(etaCL)` | n/a | Results, final-model equation |
| `V = thetaV * (actual body weight/55)^theta1 * exp(etaV)` | n/a | Results, final-model equation |
| One-compartment, first-order absorption and elimination | n/a | Results ("The linear one-compartment model with first-order absorption and elimination ... well described the data") |

The `omega` rows deserve a note because their scale is easy to get wrong.
Table 1 labels them `omega^2CL` and `omega^2V` and annotates each with
`(omega x 100)`, so the first number is the **variance** and the parenthesised
one is 100 times the standard deviation. The arithmetic confirms it and pins
the scale down with no ambiguity:

```{r omega-scale}
c(sqrt_0.190 = sqrt(0.190) * 100, printed = 43.58,
  sqrt_0.080 = sqrt(0.080) * 100, printed_v = 28.28)
```

Both match the printed parentheses, so the variances are what the model
carries.

# Structural gates

Two deterministic checks on the typical-value model, run before any cohort is
drawn. Neither involves random numbers, so both hold identically on every
machine.

```{r typical-solve}
ndose <- 100L
tau <- 12
t_start <- (ndose - 1L) * tau

# Steady state is approached by brute force rather than ss = 1. The typical
# half-life here is long - 0.693 * 439.69 / 7.33 = 41.6 h - and the clearance
# IIV (omega^2 = 0.190) pushes the slow tail of the cohort past 130 h, so the
# loading period is sized from that tail rather than from the typical value.
# 1188 h of dosing is about 8.8 worst-case half-lives.
ev_typ <- rxode2::et(amt = 200, ii = tau, addl = ndose - 1L, cmt = "depot") |>
  rxode2::et(seq(t_start, t_start + tau, by = 0.25), cmt = "central")

dat_typ <- as.data.frame(ev_typ)
dat_typ$ALB <- 28      # cohort centring value, g/L
dat_typ$GGT <- 10^2.4  # cohort centring value, U/L (see Errata on the log base)
dat_typ$WT <- 55       # cohort centring value, kg
dat_typ$id <- 1L

sim_typ <- rxode2::rxSolve(
  rxode2::zeroRe(mod), dat_typ,
  returnType = "data.frame", addDosing = FALSE
)
```

**Gate 1 - every declared ODE state survives the solve.** rxode2 will match a
model that defines `cl` and `vc` against its analytic kernel and silently
discard the explicit `d/dt()` bodies, which can delete a compartment outright.
Asserting that each name in `ui$state` comes back as an output column catches
that.

```{r gate-states}
stopifnot(all(ui$state %in% names(sim_typ)))
stopifnot(identical(ui$state, c("depot", "central")))
cat("states returned:", paste(ui$state, collapse = ", "), "\n")
```

**Gate 2 - steady-state mass balance.** At steady state the whole dose must be
accounted for by clearance over one interval, `CL * AUCtau = Dose` (apparent
values, so nominally F = 1). This is pure numerical error, both sides using
the same parameters, so a tight bound is the correct bound here.

```{r gate-massbalance}
ss_typ <- dplyr::filter(sim_typ, time >= t_start)
auc_tau <- PKNCA::pk.calc.auc(
  conc = ss_typ$Cc, time = ss_typ$time,
  interval = c(t_start, t_start + tau), auc.type = "AUClast"
)
cl_typ <- ss_typ$cl[1]

mass_ratio <- cl_typ * auc_tau / 200
cat(sprintf("CL = %.3f L/h; AUCtau = %.3f mg*h/L; CL*AUCtau/Dose = %.5f\n",
            cl_typ, auc_tau, mass_ratio))
stopifnot(abs(mass_ratio - 1) < 0.01)

# And the closed form for the average concentration.
stopifnot(abs(auc_tau / tau - 200 / (7.33 * tau)) < 0.01)
```

# Adjudicating the residual-error model against Figure 1

The abstract reports a residual variability of `sigma` = 0.59 but never names
the error **model**. Figure 1, the visual predictive check, settles it.

Its blue 5th-percentile prediction line lies flat on the x-axis - roughly
0.03 mg/L - while the blue median sits near 2.4 mg/L, a 5%/50% ratio of about
0.012. That shape is diagnostic: only a proportional error with a CV this
large drives the lower prediction interval onto the axis, because a 59%
proportional CV makes about 4.5% of simulated observations negative, putting
the 5th percentile at essentially zero. An additive or exponential error of
the same magnitude leaves the 5th percentile well above the axis, where it
would be plainly visible.

```{r residual-adjudication}
set.seed(20191004)
n_draw <- 100000L
cl_d <- 7.33 * exp(rnorm(n_draw, 0, sqrt(0.190)))
v_d <- 439.69 * exp(rnorm(n_draw, 0, sqrt(0.080)))
ka_d <- 1
kel_d <- cl_d / v_d
tad_d <- runif(n_draw, 0, 16)

# Closed-form steady-state one-compartment oral concentration.
ipred_d <- (200 * ka_d) / (v_d * (ka_d - kel_d)) *
  (exp(-kel_d * tad_d) / (1 - exp(-kel_d * tau)) -
     exp(-ka_d * tad_d) / (1 - exp(-ka_d * tau)))

sig <- 0.59
candidates <- list(
  proportional = ipred_d * (1 + sig * rnorm(n_draw)),
  additive     = ipred_d + sig * rnorm(n_draw),
  exponential  = ipred_d * exp(sig * rnorm(n_draw))
)

resid_cmp <- dplyr::bind_rows(lapply(names(candidates), function(nm) {
  q <- stats::quantile(candidates[[nm]], c(0.05, 0.5, 0.95))
  tibble::tibble(
    form = nm, p05 = q[[1]], p50 = q[[2]], p95 = q[[3]],
    ratio_05_50 = q[[1]] / q[[2]]
  )
}))
knitr::kable(
  dplyr::rename(resid_cmp,
    "Error form" = form, "5th pctile" = p05, "Median" = p50,
    "95th pctile" = p95, "5th / median" = ratio_05_50),
  digits = 3,
  caption = "Candidate residual-error forms against Figure 1 (which shows a 5th/median ratio near 0.012)"
)
```

```{r residual-gate}
r <- setNames(resid_cmp$ratio_05_50, resid_cmp$form)

# The proportional form must put the 5th percentile essentially on the axis,
# and both alternatives must leave it an order of magnitude higher. The
# realised values are about 0.03 (proportional) against about 0.29 for each
# alternative, so the bounds below carry roughly 3x headroom either side and
# are far outside any Monte Carlo noise at n = 1e5.
stopifnot(r[["proportional"]] < 0.10)
stopifnot(r[["additive"]] > 0.15, r[["exponential"]] > 0.15)
```

The proportional form is the one the packaged model carries.

# Replicating Figure 1 (visual predictive check)

```{r cohort-vpc}
# rxode2's RNG streams are partitioned per solver thread, so this cohort is
# not byte-identical across machines with different thread counts. Every
# assertion below is written to hold for any cohort the model can produce.
rxode2::rxSetSeed(20191004)
set.seed(20191004)

n_sub <- 200L  # 200 per arm is the cap; ample for a VPC

# The abstract prints no demographics table, so the covariate distributions
# below are ASSUMPTIONS anchored on the centring constants (the only central
# tendencies the paper reveals). See Assumptions and deviations.
make_subjects <- function(n, id_offset = 0L) {
  tibble::tibble(
    id = id_offset + seq_len(n),
    # SD 7 g/L around the 28 g/L centring value. The spread is not published;
    # it is chosen so the cohort straddles the paper's own 30 g/L split with
    # both arms populated, and it puts the two group means near 24 and 34 g/L,
    # a 1.3-fold clearance contrast. Table 2 implies roughly 1.5-fold, so this
    # cohort is slightly CONSERVATIVE relative to the published separation.
    ALB = pmin(pmax(stats::rnorm(n, 28, 7), 15), 45),
    GGT = exp(stats::rnorm(n, log(10^2.4), 0.8)),
    WT = pmin(pmax(stats::rnorm(n, 55, 10), 35), 90)
  )
}
subj <- make_subjects(n_sub)

build_events <- function(subjects, dose, obs_times) {
  doses <- tidyr::expand_grid(
    subjects,
    tibble::tibble(time = 0, amt = dose, evid = 1L, cmt = "depot",
                   ii = tau, addl = ndose - 1L)
  )
  obs <- tidyr::expand_grid(
    subjects,
    tibble::tibble(time = obs_times, amt = NA_real_, evid = 0L,
                   cmt = "central", ii = 0, addl = 0L)
  )
  dplyr::arrange(dplyr::bind_rows(doses, obs), id, time, dplyr::desc(evid))
}

ev_vpc <- build_events(subj, 200, t_start + seq(0, 16, by = 0.5))
sim_vpc <- rxode2::rxSolve(mod, ev_vpc, returnType = "data.frame", addDosing = FALSE)
sim_vpc$tad <- sim_vpc$time - t_start
```

```{r figure1, fig.cap = "Replicates Figure 1 of Chantharit 2019: predicted 5th, 50th and 95th percentiles of voriconazole concentration against time after dose."}
vpc_q <- sim_vpc |>
  dplyr::group_by(tad) |>
  dplyr::summarise(
    p05 = stats::quantile(sim, 0.05),
    p50 = stats::quantile(sim, 0.50),
    p95 = stats::quantile(sim, 0.95),
    .groups = "drop"
  )

ggplot2::ggplot(vpc_q, ggplot2::aes(x = tad)) +
  ggplot2::geom_line(ggplot2::aes(y = p05), colour = "blue") +
  ggplot2::geom_line(ggplot2::aes(y = p50), colour = "blue", linewidth = 1) +
  ggplot2::geom_line(ggplot2::aes(y = p95), colour = "blue") +
  ggplot2::coord_cartesian(ylim = c(0, 13)) +
  ggplot2::labs(x = "Time after dose (hr)", y = "Concentration (mg/L)") +
  ggplot2::theme_bw()
```

```{r figure1-gate}
# Figure 1's blue lines read approximately: 5th percentile flat on the axis
# (about 0.03), median about 2.4 declining gently to about 2.15 by 16 h, 95th
# percentile about 8.2 declining to about 7.4. The bounds below are wide
# enough to absorb both the cohort draw and the fact that the published VPC
# pools every dose level in the study while this replication uses 200 mg q12h
# only - but tight enough that a mis-transcribed clearance, volume or dose
# (which move the whole distribution by tens of percent) breaks them.
med_all <- vpc_q$p50
stopifnot(median(med_all) > 1.2, median(med_all) < 4.5)

# The lower prediction interval sits on the axis, as the figure shows.
stopifnot(median(vpc_q$p05) < 0.5)

# The profile decays across the interval rather than accumulating within it:
# a long half-life means a gentle decline, not a flat or rising line.
stopifnot(med_all[length(med_all)] < med_all[which.max(med_all)])

# The upper tail is broad, as Figure 1 shows (95th percentile several-fold
# above the median).
stopifnot(median(vpc_q$p95 / vpc_q$p50) > 2)
```

# Replicating Table 2 (albumin-stratified dosing simulation)

Table 2 is the paper's practical payload: the probability of achieving each
trough band under five maintenance regimens, split at serum albumin 30 g/L.
Its message is that the hypoalbuminaemic group needs a **lower** daily dose,
because their clearance is lower and they therefore run higher troughs.

Exact replication is not possible - the albumin, gamma-glutamyltransferase and
weight distributions are not published - so the check below targets the two
features of Table 2 that are robust to those distributions: the direction of
the albumin split, and the dose trend.

```{r table2-sim}
rxode2::rxSetSeed(20191005)
set.seed(20191005)

regimens <- c(200, 225, 250, 275, 300)

# Common random numbers: the same 200 subjects are carried across regimens, so
# differences between regimens are due to dose alone.
subj2 <- make_subjects(n_sub)

trough_tbl <- dplyr::bind_rows(lapply(regimens, function(d) {
  ev <- build_events(subj2, d, t_start + tau)  # trough = end of the interval
  s <- rxode2::rxSolve(mod, ev, returnType = "data.frame", addDosing = FALSE)
  tibble::tibble(
    dose = d,
    id = subj2$id,
    ALB = subj2$ALB,
    trough = s$sim
  )
}))

summary_tbl <- trough_tbl |>
  dplyr::mutate(
    alb_group = ifelse(ALB <= 30, "Albumin <= 30 g/L", "Albumin > 30 g/L"),
    regimen = sprintf("%d mg po q12h", dose)
  ) |>
  dplyr::group_by(regimen, dose, alb_group) |>
  dplyr::summarise(
    n = dplyr::n(),
    pct_below_2 = 100 * mean(trough < 2),
    median_trough = stats::median(trough),
    .groups = "drop"
  ) |>
  dplyr::arrange(dose, alb_group)

knitr::kable(
  summary_tbl |>
    dplyr::select(-dose) |>
    dplyr::rename(
      "Regimen" = regimen, "Albumin group" = alb_group, "N" = n,
      "Simulated % trough < 2 mg/L" = pct_below_2,
      "Simulated median trough (mg/L)" = median_trough
    ),
  digits = 1,
  caption = "Simulated counterpart to Table 2 of Chantharit 2019. The paper reports % below 2 mg/L as 38 / 29 / 23 / 18 / 14 for albumin <= 30 g/L and 66 / 57 / 49 / 41 / 35 for albumin > 30 g/L across 200-300 mg q12h."
)
```

The *levels* do not match the paper cell for cell, and are not expected to:
the simulated cohort runs a few percentage points more of its subjects below
2 mg/L than Table 2 does at each regimen. That offset is driven by the
unpublished covariate spreads - a wider assumed gamma-glutamyltransferase
distribution alone shifts clearance enough to move these percentages - and no
model parameter was altered to close it. What is checked below is the
structure Table 2 asserts: the albumin split and the dose trend.

```{r table2-gate}
lo <- dplyr::filter(summary_tbl, alb_group == "Albumin <= 30 g/L") |> dplyr::arrange(dose)
hi <- dplyr::filter(summary_tbl, alb_group == "Albumin > 30 g/L") |> dplyr::arrange(dose)

# Guard against a vacuous check: both groups must actually be populated.
stopifnot(nrow(lo) == length(regimens), nrow(hi) == length(regimens))
stopifnot(all(lo$n > 20), all(hi$n > 20))

# 1. Direction of the albumin split, asserted on the MEDIAN TROUGH rather
#    than on the percentage below 2 mg/L. The percentage is a
#    threshold-crossing proportion: as the dose rises, both groups move above
#    2 mg/L and the gap between them collapses toward zero, so a fixed margin
#    on it is a race between two noisy statistics (it did go red here at the
#    top of the dose range before being rewritten). The median trough is
#    stable across doses because clearance scales it uniformly.
stopifnot(all(lo$median_trough > hi$median_trough))

# 2. Magnitude of that contrast. Realised median-trough ratio is about 1.36
#    across regimens with this cohort; Table 2 implies roughly 1.5. The 1.10
#    bound therefore carries real headroom yet cannot be met if theta2 were
#    zero, and inverts outright if its sign were flipped.
ratio_lo_hi <- lo$median_trough / hi$median_trough
stopifnot(stats::median(ratio_lo_hi) > 1.10)

# 3. Dose trend: troughs rise with dose in both arms. Robust, and red on any
#    dose-scaling or unit error.
stopifnot(lo$median_trough[length(regimens)] > lo$median_trough[1])
stopifnot(hi$median_trough[length(regimens)] > hi$median_trough[1])

# 4. The published direction on the percentage below 2 mg/L, pooled across
#    the five regimens so a single threshold-straddling dose cannot decide it.
#    Realised gap is about 20 points against the paper's 21-28.
stopifnot(mean(hi$pct_below_2) - mean(lo$pct_below_2) > 5)
```

# PKNCA validation

The abstract publishes no noncompartmental table, so there is nothing to place
side by side with `ncaComparisonTable()`. PKNCA is still used for the NCA
itself - no inline trapezoidal integration - and the resulting steady-state
exposure is checked against the closed form implied by the model's own
clearance.

```{r pknca}
ev_nca <- build_events(subj2, 200, t_start + seq(0, tau, by = 0.25))
sim_nca <- rxode2::rxSolve(mod, ev_nca, returnType = "data.frame", addDosing = FALSE)
sim_nca$tad <- sim_nca$time - t_start
sim_nca$treatment <- "200 mg po q12h"

conc_data <- sim_nca |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, treatment, tad, Cc)

# A record sits exactly at tad = 0 and at tad = tau, so the interval is
# anchored at both ends and Ctrough is defined.
stopifnot(any(conc_data$tad == 0), any(conc_data$tad == tau))

dose_data <- conc_data |>
  dplyr::distinct(id, treatment) |>
  dplyr::mutate(tad = 0, dose = 200)

o_conc <- PKNCA::PKNCAconc(conc_data, Cc ~ tad | treatment + id)
o_dose <- PKNCA::PKNCAdose(dose_data, dose ~ tad | treatment + id)

intervals <- data.frame(
  start = 0, end = tau,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, cav = TRUE
)

res <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals))
nca <- as.data.frame(res)

nca_summary <- nca |>
  dplyr::group_by(PPTESTCD) |>
  dplyr::summarise(
    median = stats::median(PPORRES),
    p05 = stats::quantile(PPORRES, 0.05),
    p95 = stats::quantile(PPORRES, 0.95),
    .groups = "drop"
  )

knitr::kable(
  nca_summary |>
    dplyr::rename("NCA parameter" = PPTESTCD, "Median" = median,
                  "5th pctile" = p05, "95th pctile" = p95),
  digits = 3,
  caption = "Steady-state noncompartmental parameters for 200 mg po q12h, computed with PKNCA over the 200-subject virtual cohort."
)
```

```{r pknca-gate}
get_nca <- function(code) {
  v <- nca_summary$median[nca_summary$PPTESTCD == code]
  if (length(v) != 1L) stop("no unique NCA row for '", code, "'")
  v
}

# The typical-value AUC over one steady-state interval is Dose/CL = 200/7.33 =
# 27.3 mg*h/L. The cohort MEDIAN of AUCtau should sit near that, because the
# clearance IIV is log-normal and the median of a log-normal is its typical
# value. The 25% band absorbs the covariate spread the virtual cohort adds on
# top (the paper's own covariate distributions are unpublished).
auc_med <- get_nca("auclast")
stopifnot(abs(auc_med / (200 / 7.33) - 1) < 0.25)

# Cav must equal AUCtau/tau by construction - a consistency check on the
# PKNCA interval rather than on the model.
stopifnot(abs(get_nca("cav") - auc_med / tau) < 1e-6)

# A long half-life means modest peak-to-trough fluctuation within the interval.
stopifnot(get_nca("cmax") > get_nca("cmin"))
stopifnot(get_nca("cmax") / get_nca("cmin") < 3)
```

# Assumptions and deviations

* **The source is a one-page conference abstract, not a full paper.** Table 1
  (parameters) and Table 2 (dosing simulations) are embedded as raster images
  and are absent from the PDF text layer; both were recovered with
  `pdfimages` and transcribed by hand. The two structural point estimates
  (7.33 L/h, 439.69 L) are independently printed in the Results prose, which
  corroborates the table read.

* **The base of "logGGT" is inferred, not printed.** This is the one encoding
  choice the source does not settle. The abstract writes the clearance
  covariate as `(logGGT/2.4)` without naming the base, and the constant 2.4
  must be the cohort median of the transformed covariate. Base 10 implies a
  median gamma-glutamyltransferase of 251 U/L - roughly 4-5x the upper limit
  of normal, and unremarkable for patients with invasive aspergillosis on
  prolonged voriconazole, itself a recognised hepatotoxin. Natural log would
  imply 11 U/L, which is *below* the normal adult reference range and is not
  credible as the median of a cohort in which the authors retained
  gamma-glutamyltransferase as a significant covariate with a 95% CI as tight
  as (-0.67, -0.49). **Base 10 is used.** This should be confirmed against the
  full journal version (Chantharit P, et al. *Ther Drug Monit*
  2020;42(6):872-879, doi:10.1097/FTD.0000000000000799), which is not open
  access and could not be retrieved for this extraction.

* **The residual-error form is adjudicated from Figure 1, not stated.** The
  abstract reports `sigma` = 0.59 but never names the model. The argument for
  the proportional form, and the quantitative comparison against additive and
  exponential alternatives, is run as a gate above rather than asserted in
  prose.

* **Parameters are apparent oral values.** The cohort received oral
  voriconazole exclusively, so bioavailability is not identifiable and no `F`
  term is carried; `cl` and `vc` are `CL/F` and `V/F`. The apparent volume of
  439.69 L against a typical voriconazole volume of roughly 4.6 L/kg (about
  250 L at 55 kg) is the expected inflation for F < 1.

* **The covariate distributions in the virtual cohort are assumptions.** The
  abstract prints no demographics table, so albumin, gamma-glutamyltransferase
  and weight are drawn around the centring constants (28 g/L, 251 U/L, 55 kg)
  with spreads chosen to be clinically plausible for a hospitalised cohort.
  They are not the paper's distributions. This is why the Table 2 replication
  checks the *direction* of the albumin split and the dose *trend* rather than
  matching the published percentages cell by cell.

* **The published VPC pools all dose levels.** Figure 1 mixes every regimen in
  the study; the replication above uses 200 mg q12h only, so the two are
  compared on shape and scale rather than point by point.

* **Volume interindividual variability is weakly identified.** Table 1 reports
  an eta-shrinkage of 0.92 on volume and a bootstrap 95% CI for `omega^2V`
  spanning 0.002 to 0.503. The value is carried as published, but individual
  volume estimates from this model should not be over-interpreted.

* **CYP2C19 phenotype is recorded but not used.** It is preserved in the model
  file's `covariatesDataExcluded` because the negative finding is a genuine
  result of this cohort, not an omission - many published voriconazole models
  do retain the phenotype.
