---
title: "Systemic agents for psoriasis: PASI response (Checchio 2017)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Systemic agents for psoriasis: PASI response (Checchio 2017)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Checchio 2017 is a model-based meta-analysis (MBMA) of systemic antipsoriatic
agents. It develops **two independent models** from one curated database, and
this package ships each as its own model file because the source fitted them
separately, on different datasets, in different software:

* `Checchio_2017_psoriasis_pasi75_longitudinal_mbma` -- the **longitudinal**
  model, describing how the PASI75 responder rate rises over time for each of
  13 drugs (NONMEM 7.3, 57 studies).
* `Checchio_2017_psoriasis_pasi_landmark_mbma` -- the **landmark** model,
  describing the Week-12 dose-response across all four PASI endpoints
  (PASI50/75/90/100) for 16 drug arms (S-PLUS 8.0.4, 71 studies).

* Article: <https://doi.org/10.1002/cpt.732> (PMC5697570)
* Supplement: `CPT-102-1006-s001.docx`, retrieved from the EuropePMC
  `supplementaryFiles` endpoint for PMC5697570. It carries Supplementary
  Tables S1.1, S1.2 and 2, which hold **every** parameter estimate; none is in
  the main text.

```{r models}
modLong <- readModelDb("Checchio_2017_psoriasis_pasi75_longitudinal_mbma")
modLand <- readModelDb("Checchio_2017_psoriasis_pasi_landmark_mbma")
uiLong <- rxode2::rxode2(modLong)
uiLand <- rxode2::rxode2(modLand)
```

**A note on reading the source.** Every display equation in the published PDF
is a rasterised image. None of the thirteen survives plain-text extraction of
the PDF, and a reading that worked from such an extraction alone would have no
structural model at all. All thirteen were recovered with
`pdftotext -layout`, which reproduces the mathematical text (mangled in
places, but unambiguously) directly from the PDF's font layer.

# Population

Both models operate at the **study-arm** level: one observation is one trial
arm's responder rate, not one patient's outcome. The random effects are
therefore *between-study*, and neither model may be used to simulate
individual patients.

The database came from an Ovid Medline / Summary-Basis-of-Approval /
European-Public-Assessment-Report search covering 1998-2015 and following the
Cochrane approach, plus an internal Pfizer tofacitinib database (all
contributing tofacitinib studies are themselves published; the internal source
was used to preserve numerical accuracy). The search yielded 912 abstracts, of
which 151 studies were screened in, 71 survived into the landmark analysis and
57 into the longitudinal analysis.

Body weight is the only covariate, entered as a study-arm mean. The dataset
median is approximately **90 kg**, which is the anchor in both covariate
equations and the weight at which every published prediction is made. Where an
arm did not report weight (10-15% of studies) the source set it to 90 so the
covariate term vanishes; downstream code should do the same.

```{r population}
pop <- modLong()$population
tibble::tibble(
  Field = c("Species", "Studies (longitudinal)", "Studies (landmark)", "Disease state", "Weight anchor", "Timepoints"),
  Value = c(
    pop$species, "57", "71", pop$disease_state, pop$weight_range,
    "longitudinal: repeated arm read-outs, primarily Weeks 4 and 12; landmark: a single Week-12 read-out (range 10-24 weeks)"
  )
) |>
  knitr::kable()
```

# Source trace

Every model equation and every `ini()` value, with where it came from. The
in-file comments carry the same trace per line; this table is the single place
a reviewer can audit provenance.

```{r source-trace-equations}
tibble::tribble(
  ~Component, ~Source,
  "Pr(PASI75) = inverse-logit(f0 + fdrug)", "Equations 1-2, Methods 'Longitudinal model'",
  "f0 = A + B * (1 - exp(-kpbo * t)) * exp(eta)", "Equation 3",
  "fdrug = Emax_d * (1 - exp(-kdrug_d * t)) * Dose / (Dose + ED50_d)", "Equation 4 (Hill exponent tested, not retained; see Errata)",
  "P(event) = inverse-logit(E0_i + Edrug + eta_i,k)", "Equation 5, Methods 'Landmark model'",
  "E0_i = trial_i + I1*PASI50 + I2*PASI90 + I3*PASI100", "Equation 6",
  "em = em_d * DOSE^gamma / (exp(ED50_d)^gamma + DOSE^gamma)", "Equation 7 (traditional oral agents: em = em_d, a step)",
  "Edrug = em * (1 + I5*PASI50 + I6*PASI90 + I7*PASI100)", "Equation 8",
  "Longitudinal weight effect: (WEIG / 90)^theta", "Equation 9, Methods 'Covariate model'",
  "Landmark weight effect: (WEIG - 90) * I4", "Equation 10",
  "Residual: response = Pr + W * eps, W = sqrt(Pr*(1-Pr)/N)", "Equations 11-12, Methods 'Residual error model'",
  "Longitudinal third variance level: Pr + W * (eps + eps_corr)", "Equation 13"
) |>
  knitr::kable(caption = "Structural, covariate and residual equations.")
```

```{r source-trace-params}
tibble::tribble(
  ~Block, ~Parameters, ~Source,
  "Longitudinal, common", "A, B, kpbo, weight effects on B and kdrug, omega, sigma, sigma_corr", "Supplementary Table S1.2",
  "Longitudinal, per drug", "Emax, ED50, kdrug for 13 drugs", "Supplementary Table S1.1",
  "Landmark, placebo", "I1, I2, I3, I4", "Supplementary Table 2, e_o block",
  "Landmark, Emax", "bio and dmard intercepts, vitamin A analog, four class offsets", "Supplementary Table 2, em block",
  "Landmark, Hill", "nt intercept and IL-17 offset (log scale, per footnote a)", "Supplementary Table 2, nt block",
  "Landmark, ED50", "14 log-scale rows including alefacept IV and brodalumab Q4W offsets", "Supplementary Table 2, ed block",
  "Landmark, endpoint scaling", "em50, em90, em100", "Supplementary Table 2, scaling block",
  "Longitudinal, PASI90 scaling", "i_pbo_pasi90, e_drug_pasi90 (imported, hence fixed())", "Supplementary Table 2, via Methods 'External validation'",
  "Landmark, trial intercept", "e0_trial -- NOT PRINTED; back-solved from Table 2 below", "derived (see 'Recovering the two unprinted values')",
  "Landmark, briakinumab ED50", "led50_briakinumab -- NOT PRINTED; back-solved from Table 2 below", "derived (see 'Recovering the two unprinted values')"
) |>
  knitr::kable(caption = "Parameter provenance by block.")
```

# Study arms

Both models take dose through one `CONMED_<drug>_DOSE` covariate column per
drug, and zero in every other column. The arm table below encodes the
"clinical dose" column of the source's Tables 1 and 2. **The dose is per
administration**, not a daily or weekly total -- see the Errata for the
arithmetic that settles this.

```{r arms}
doseCols <- c(
  "CONMED_ADALIMUMAB_DOSE", "CONMED_CERTOLIZUMAB_DOSE", "CONMED_ETANERCEPT_DOSE",
  "CONMED_INFLIXIMAB_DOSE", "CONMED_BRIAKINUMAB_DOSE", "CONMED_USTEKINUMAB_DOSE",
  "CONMED_BRODALUMAB_DOSE", "CONMED_IXEKIZUMAB_DOSE", "CONMED_SECUKINUMAB_DOSE",
  "CONMED_ALEFACEPT_DOSE", "CONMED_APREMILAST_DOSE", "CONMED_MTX_DOSE",
  "CONMED_TOFACITINIB_DOSE", "CONMED_BARICITINIB_DOSE", "CONMED_CSA_DOSE",
  "CONMED_ACITRETIN_DOSE"
)

arms <- tibble::tribble(
  ~arm,            ~regimen,        ~col,                       ~dose, ~routeIv, ~inLong,
  "Placebo",       "--",            NA_character_,                  0,        0,    TRUE,
  "Adalimumab",    "40 mg Q2W",     "CONMED_ADALIMUMAB_DOSE",      40,        0,    TRUE,
  "Certolizumab",  "200 mg Q2W",    "CONMED_CERTOLIZUMAB_DOSE",   200,        0,    TRUE,
  "Etanercept 25", "25 mg BIW",     "CONMED_ETANERCEPT_DOSE",      25,        0,    TRUE,
  "Etanercept 50", "50 mg BIW",     "CONMED_ETANERCEPT_DOSE",      50,        0,    TRUE,
  "Infliximab",    "5 mg/kg Q8W",   "CONMED_INFLIXIMAB_DOSE",       5,        0,    TRUE,
  "Brodalumab",    "210 mg Q2W",    "CONMED_BRODALUMAB_DOSE",     210,        0,    TRUE,
  "Ixekizumab",    "80 mg Q4W",     "CONMED_IXEKIZUMAB_DOSE",      80,        0,    TRUE,
  "Secukinumab",   "150 mg QM",     "CONMED_SECUKINUMAB_DOSE",    150,        0,    TRUE,
  "Briakinumab",   "100 mg Q4W",    "CONMED_BRIAKINUMAB_DOSE",    100,        0,    TRUE,
  "Ustekinumab",   "45 mg Q12W",    "CONMED_USTEKINUMAB_DOSE",     45,        0,    TRUE,
  "Methotrexate",  "18 mg QW",      "CONMED_MTX_DOSE",             18,        0,    TRUE,
  "Tofacitinib 5", "5 mg BID",      "CONMED_TOFACITINIB_DOSE",      5,        0,    TRUE,
  "Tofacitinib 10", "10 mg BID",    "CONMED_TOFACITINIB_DOSE",     10,        0,    TRUE,
  "Alefacept",     "10 mg QW",      "CONMED_ALEFACEPT_DOSE",       10,        1,    TRUE,
  "Apremilast",    "30 mg BID",     "CONMED_APREMILAST_DOSE",      30,        0,    TRUE,
  "Baricitinib",   "10 mg QD",      "CONMED_BARICITINIB_DOSE",     10,        0,   FALSE,
  "Acitretin",     "30 mg QD",      "CONMED_ACITRETIN_DOSE",       30,        0,   FALSE,
  "Ciclosporin",   "2.5-5 mg/kg/d", "CONMED_CSA_DOSE",              5,        0,   FALSE
)

#' Build an rxode2 input frame: one id per arm, observations at `times`.
makeArmData <- function(armTbl, times, wt = 90, nArm = 100) {
  do.call(rbind, lapply(seq_len(nrow(armTbl)), function(i) {
    d <- data.frame(id = i, time = times, WT = wt, N_ARM = nArm,
                    ROUTE_IV = armTbl$routeIv[i], REGI_Q4W = 0)
    for (cc in doseCols) d[[cc]] <- 0
    if (!is.na(armTbl$col[i])) d[[armTbl$col[i]]] <- armTbl$dose[i]
    d
  }))
}

arms |>
  select(Arm = arm, `Clinical dose` = regimen, `In longitudinal model` = inLong) |>
  knitr::kable(caption = "Study arms, from the 'Clinical dose' columns of Checchio 2017 Tables 1 and 2. Ciclosporin, acitretin and baricitinib are landmark-only: Results 'Available data' excluded them from the longitudinal analysis for insufficient longitudinal data.")
```

# Longitudinal model

## Reproducing Table 1

Table 1 of the source reports, for a typical 90 kg patient at each drug's
clinical dose, the predicted PASI75 responder rate at Weeks 4 and 12 and the
onset metrics ET50 and ET90 (the times to reach 50% and 90% of the maximal
treatment effect). These are **derived quantities, not parameter estimates**,
so reproducing them is an independent check on the whole encoding.

```{r long-sim}
armsLong <- arms |> filter(inLong)
tGrid <- c(seq(0, 30, by = 0.05), 500) # 500 weeks stands in for the plateau
solLong <- rxode2::rxSolve(
  rxode2::zeroRe(uiLong), makeArmData(armsLong, tGrid),
  returnType = "data.frame"
) |>
  mutate(arm = armsLong$arm[id])

#' Time at which the PASI75 response first reaches `frac` of its plateau.
etx <- function(d, frac) {
  plateau <- d$prob_pasi75[d$time == 500]
  d <- d[d$time <= 30, ]
  target <- frac * plateau
  if (max(d$prob_pasi75) < target) return(NA_real_)
  stats::approx(d$prob_pasi75, d$time, xout = target)$y
}

longMod <- solLong |>
  group_by(arm) |>
  summarise(
    wk4  = 100 * prob_pasi75[time == 4],
    wk12 = 100 * prob_pasi75[time == 12],
    et50 = etx(pick(everything()), 0.5),
    et90 = etx(pick(everything()), 0.9),
    .groups = "drop"
  )
```

```{r long-table1}
# Checchio 2017 Table 1, transcribed. ET90 for alefacept is reported as
# "> 24 (23.5, NA)" because its data stop at 24 weeks; treated as NA here.
table1 <- tibble::tribble(
  ~arm,             ~wk4Pub, ~wk12Pub, ~et50Pub, ~et90Pub,
  "Adalimumab",       15.60,     65.9,      6.0,     11.7,
  "Certolizumab",     38.80,     74.5,      4.2,      9.9,
  "Etanercept 25",     4.64,     32.0,      8.2,     16.4,
  "Etanercept 50",     6.87,     48.7,      7.7,     14.6,
  "Infliximab",       32.70,     75.3,      4.7,     10.0,
  "Brodalumab",       46.90,     80.2,      3.9,      9.2,
  "Ixekizumab",       48.80,     81.7,      3.7,      8.9,
  "Secukinumab",      21.50,     69.2,      5.5,     11.0,
  "Briakinumab",      19.00,     78.4,      5.8,     10.9,
  "Ustekinumab",      10.20,     65.2,      7.1,     13.3,
  "Methotrexate",      3.38,     28.5,      9.2,     18.7,
  "Tofacitinib 5",     9.21,     40.8,      6.9,     14.3,
  "Tofacitinib 10",   16.90,     60.7,      6.0,     12.2,
  "Alefacept",         1.99,     15.6,     11.1, NA_real_,
  "Apremilast",        5.66,     28.5,      7.4,     15.1
)

cmpLong <- table1 |>
  left_join(longMod, by = "arm") |>
  mutate(
    dWk4 = wk4 - wk4Pub, dWk12 = wk12 - wk12Pub,
    dEt50 = et50 - et50Pub, dEt90 = et90 - et90Pub
  )

cmpLong |>
  transmute(
    Arm = arm,
    `Week 4 published (%)` = round(wk4Pub, 2), `Week 4 model (%)` = round(wk4, 2),
    `Week 12 published (%)` = round(wk12Pub, 1), `Week 12 model (%)` = round(wk12, 1),
    `ET50 published (wk)` = et50Pub, `ET50 model (wk)` = round(et50, 1)
  ) |>
  knitr::kable(caption = "Longitudinal model against Checchio 2017 Table 1 (typical 90 kg arm).")
```

```{r long-gate}
# Deterministic comparison: both sides are typical-value arithmetic at the
# identical published parameter values, so the only difference is the source's
# 3-significant-figure rounding. A tight all() bound is therefore the correct
# gate here -- there is no simulated cohort whose extreme could drift.
stopifnot(
  # Placebo is essentially zero at randomisation, as PASI75 must be.
  solLong$prob_pasi75[solLong$time == 0] < 1e-3,
  # Every published Week-4 and Week-12 rate reproduced.
  max(abs(cmpLong$dWk4)) < 1.0,
  max(abs(cmpLong$dWk12)) < 1.5,
  # Every published onset time reproduced.
  max(abs(cmpLong$dEt50)) < 0.5,
  # Alefacept's ET90 is reported only as "> 24"; check that, not a value.
  cmpLong$et90[cmpLong$arm == "Alefacept"] > 24,
  max(abs(cmpLong$dEt90[cmpLong$arm != "Alefacept"])) < 1.5
)
cat(sprintf(
  "Week 12 PASI75: max |deviation| = %.2f percentage points across %d arms\nET50:           max |deviation| = %.2f weeks\n",
  max(abs(cmpLong$dWk12)), nrow(cmpLong), max(abs(cmpLong$dEt50))
))
```

ET90 runs systematically about one week short of the published value for the
slower drugs (methotrexate 17.4 vs 18.7, etanercept 25 mg 15.2 vs 16.4). ET50,
which is the onset metric the source actually discusses, matches everywhere to
within 0.3 weeks. The likely cause is that the source computed ET90 against a
plateau evaluated at a finite horizon rather than at infinity; it is recorded
in the Errata rather than papered over.

## Figure 2: PASI75 time course

```{r fig2, fig.cap = "Replicates Figure 2 of Checchio 2017: model-predicted PASI75 time course by drug at the clinical dose, for a typical 90 kg arm."}
solLong |>
  filter(time <= 24) |>
  mutate(arm = factor(arm, levels = armsLong$arm)) |>
  ggplot(aes(time, 100 * prob_pasi75, colour = arm)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time (weeks)", y = "PASI75 responders (%)", colour = NULL) +
  coord_cartesian(ylim = c(0, 100)) +
  theme_bw() +
  theme(legend.position = "right", legend.key.height = unit(0.8, "lines"))
```

## The body weight effect

The Discussion states that "heavier patients tend to achieve lower efficacy and
may also experience slower onset of effect compared with lighter patients".
Both weight exponents are negative, so both claims should hold. Every published
prediction is made at exactly 90 kg, where both covariate terms collapse to 1,
so this direction check is the only available test of the covariate
orientation -- and it is a real test, because a transcription sign error would
flip it.

```{r weight-effect}
wtArms <- arms |> filter(arm %in% c("Placebo", "Etanercept 50", "Ustekinumab"))
wtSol <- do.call(rbind, lapply(c(60, 90, 120), function(w) {
  rxode2::rxSolve(rxode2::zeroRe(uiLong), makeArmData(wtArms, tGrid, wt = w),
                  returnType = "data.frame") |>
    mutate(arm = wtArms$arm[id], WTkg = w)
}))

wtSummary <- wtSol |>
  group_by(arm, WTkg) |>
  summarise(wk12 = 100 * prob_pasi75[time == 12],
            et50 = etx(pick(everything()), 0.5), .groups = "drop")

wtSummary |>
  transmute(Arm = arm, `Weight (kg)` = WTkg,
            `Week 12 PASI75 (%)` = round(wk12, 1), `ET50 (wk)` = round(et50, 1)) |>
  knitr::kable(caption = "Body weight effect. Heavier arms respond less and, for drug arms, more slowly.")

stopifnot(
  # Efficacy falls monotonically with weight in every arm.
  all(wtSummary |> group_by(arm) |> arrange(WTkg) |>
        summarise(mono = all(diff(wk12) < 0), .groups = "drop") |> pull(mono)),
  # Onset slows with weight in the DRUG arms. The placebo arm carries no
  # kdrug, so its ET50 must move only through the placebo plateau shift and
  # must NOT slow -- a useful discriminator that the exponent was placed on
  # kdrug and not on kpbo.
  all(wtSummary |> filter(arm != "Placebo") |> group_by(arm) |> arrange(WTkg) |>
        summarise(mono = all(diff(et50) > 0), .groups = "drop") |> pull(mono))
)
```

# Landmark model

## Recovering the two unprinted values

The landmark model is fully tabulated in Supplementary Table 2 **except** for
two quantities:

1. **The typical study placebo intercept** (`trial_i` of Equation 6). The
   source estimated one per trial -- 71 of them -- and prints none; it says
   only that they "were shown to be normally distributed around a single
   value (data not shown)".
2. **The briakinumab ED50.** Briakinumab is predicted in Table 2 and is
   modelled by Equation 7's sigmoidal branch like every other biologic, but
   the `ed` block of Supplementary Table 2 has no briakinumab row. The
   omission is in the source, not in the conversion: the supplement's raw
   WordprocessingML was enumerated row by row and the block contains 14 rows,
   none of them briakinumab.

Both are recovered by **inverting the source's own published predictions
through the source's own published equations**, with every other parameter
fixed at its printed value. The fit below is re-run from scratch so it can be
audited, and its result is checked against the values shipped in `ini()`.

```{r backsolve}
expit <- function(x) 1 / (1 + exp(-x))

table2 <- tibble::tribble(
  ~arm,             ~p75Pub, ~p90Pub,
  "Adalimumab",        64.9,   37.90,
  "Certolizumab",      73.5,   48.20,
  "Etanercept 25",     37.4,   16.10,
  "Etanercept 50",     54.0,   27.60,
  "Infliximab",        77.0,   53.00,
  "Brodalumab",        81.2,   59.30,
  "Ixekizumab",        88.2,   71.70,
  "Secukinumab",       75.7,   51.10,
  "Briakinumab",       80.8,   58.40,
  "Ustekinumab",       70.3,   43.60,
  "Methotrexate",      36.4,   15.50,
  "Tofacitinib 5",     35.2,   14.80,
  "Tofacitinib 10",    53.8,   27.50,
  "Baricitinib",       33.2,   13.60,
  "Acitretin",         25.0,    9.57,
  "Alefacept",         22.7,    8.39,
  "Apremilast",        26.8,   10.30,
  "Ciclosporin",       46.7,   22.10
)

#' Predicted Week-12 PASI75 and PASI90 for every arm, given a trial intercept
#' and a briakinumab log ED50, evaluated through the packaged model itself.
predLandmark <- function(e0Trial, led50Bria) {
  landArms <- arms |> filter(arm != "Placebo")
  # Overridden through rxSolve(params = ), NOT by assigning into ui$theta --
  # element assignment on an rxUi theta is silently ignored.
  rxode2::rxSolve(
    uiLand, makeArmData(landArms, c(0, 12)),
    params = c(e0_trial = e0Trial, led50_briakinumab = led50Bria),
    returnType = "data.frame"
  ) |>
    filter(time == 12) |>
    transmute(arm = landArms$arm[id], p75 = 100 * prob_pasi75, p90 = 100 * prob_pasi90)
}

objective <- function(p) {
  pr <- predLandmark(p[1], p[2]) |> left_join(table2, by = "arm")
  sum((pr$p75 - pr$p75Pub)^2 + (pr$p90 - pr$p90Pub)^2)
}
fit <- stats::optim(c(-2.9, 2), objective, method = "Nelder-Mead")

tibble::tibble(
  Quantity = c("Typical study placebo intercept (e0_trial)",
               "Briakinumab log ED50 (led50_briakinumab)"),
  `Re-derived here` = round(fit$par, 4),
  `Shipped in ini()` = c(uiLand$theta[["e0_trial"]], uiLand$theta[["led50_briakinumab"]])
) |>
  knitr::kable(caption = "The two unprinted values, re-derived from Checchio 2017 Table 2.")

stopifnot(
  # The shipped values must be what this fit recovers.
  abs(fit$par[1] - uiLand$theta[["e0_trial"]]) < 0.01,
  abs(fit$par[2] - uiLand$theta[["led50_briakinumab"]]) < 0.05
)
```

The recovered placebo intercept implies a typical Week-12 placebo PASI75 of
`r sprintf("%.1f%%", 100 * expit(uiLand$theta[["e0_trial"]]))` and a typical
placebo PASI90 of
`r sprintf("%.1f%%", 100 * expit(uiLand$theta[["e0_trial"]] + uiLand$theta[["i_pbo_pasi90"]]))`.
Table 2's own "difference from placebo" columns imply 5.4% and 1.6%
respectively (adalimumab: 64.9 - 59.5 and 37.9 - 36.3), so the inversion lands
where the source's arithmetic says it should. That cross-check uses a *different*
part of Table 2 from the one the fit consumed, which is why it is worth stating
separately.

## Reproducing Table 2

```{r land-table2}
landArms <- arms |> filter(arm != "Placebo")
solLand <- rxode2::rxSolve(uiLand, makeArmData(landArms, c(0, 12)),
                           returnType = "data.frame") |>
  filter(time == 12) |>
  mutate(arm = landArms$arm[id])

cmpLand <- table2 |>
  left_join(solLand |> transmute(arm, p75 = 100 * prob_pasi75, p90 = 100 * prob_pasi90),
            by = "arm") |>
  mutate(d75 = p75 - p75Pub, d90 = p90 - p90Pub)

cmpLand |>
  transmute(
    Arm = arm,
    `PASI75 published (%)` = round(p75Pub, 1), `PASI75 model (%)` = round(p75, 1),
    `PASI90 published (%)` = round(p90Pub, 2), `PASI90 model (%)` = round(p90, 2),
    `PASI75 difference (pp)` = round(d75, 1)
  ) |>
  knitr::kable(caption = "Landmark model against Checchio 2017 Table 2 (Week 12, typical 90 kg arm).")
```

```{r land-gate}
stopifnot(
  # The bulk of the table must be reproduced tightly.
  stats::median(abs(cmpLand$d75)) < 1.0,
  stats::median(abs(cmpLand$d90)) < 1.0,
  # Two arms sit further out (see below); nothing may exceed 5 points.
  max(abs(cmpLand$d75)) < 5.0,
  max(abs(cmpLand$d90)) < 5.0
)
cat(sprintf(
  "PASI75: median |deviation| = %.2f pp, max = %.2f pp (%s)\nPASI90: median |deviation| = %.2f pp, max = %.2f pp\n",
  stats::median(abs(cmpLand$d75)), max(abs(cmpLand$d75)),
  cmpLand$arm[which.max(abs(cmpLand$d75))],
  stats::median(abs(cmpLand$d90)), max(abs(cmpLand$d90))
))
```

Sixteen of the eighteen arms are reproduced to within 1.5 percentage points.
Brodalumab (+2.7) and ustekinumab (-2.4) sit further out, and the direction is
informative: Table 2's entries are **means over 1,000 simulations that include
the between-study random effect**, whereas the values here are typical-value
predictions. Averaging `inverse-logit(x + eta)` pulls extreme probabilities
toward the middle, which is exactly the sign of the brodalumab residual at
84%. The random-effect variances are not published for the landmark model (see
Errata), so the shift cannot be reproduced rather than merely explained.

## Structural admissibility

The four endpoint probabilities must be strictly ordered for every arm --
PASI50 is a laxer threshold than PASI75, which is laxer than PASI90, which is
laxer than PASI100. Nothing in the fitting enforces this; it emerges from the
signs of the three placebo offsets and the three scaling factors, so it is a
genuine check that they were transcribed correctly.

```{r admissibility}
ord <- solLand |>
  transmute(arm, p50 = prob_pasi50, p75 = prob_pasi75,
            p90 = prob_pasi90, p100 = prob_pasi100)
stopifnot(
  all(ord$p50 > ord$p75), all(ord$p75 > ord$p90), all(ord$p90 > ord$p100),
  all(ord$p50 < 1), all(ord$p100 > 0)
)
cat("Endpoint ordering PASI50 > PASI75 > PASI90 > PASI100 holds for all",
    nrow(ord), "arms.\n")
```

## Figure 3: Week-12 dose-response

```{r fig3, fig.cap = "Replicates Figure 3 of Checchio 2017: model-predicted Week-12 PASI50/75/90/100 dose-response for four representative drugs. Dose 0 is the placebo arm of each study."}
drugGrid <- tibble::tribble(
  ~arm,          ~col,                      ~maxDose,
  "Adalimumab",  "CONMED_ADALIMUMAB_DOSE",       160,
  "Ixekizumab",  "CONMED_IXEKIZUMAB_DOSE",       160,
  "Ustekinumab", "CONMED_USTEKINUMAB_DOSE",      180,
  "Tofacitinib", "CONMED_TOFACITINIB_DOSE",       20
)
drGrid <- do.call(rbind, lapply(seq_len(nrow(drugGrid)), function(i) {
  data.frame(arm = drugGrid$arm[i], col = drugGrid$col[i],
             dose = seq(0, drugGrid$maxDose[i], length.out = 60),
             routeIv = 0, regimen = NA_character_, inLong = TRUE)
}))
solDr <- rxode2::rxSolve(uiLand, makeArmData(drGrid, c(0, 12)),
                         returnType = "data.frame") |>
  filter(time == 12) |>
  mutate(arm = drGrid$arm[id], dose = drGrid$dose[id])

solDr |>
  select(arm, dose, PASI50 = prob_pasi50, PASI75 = prob_pasi75,
         PASI90 = prob_pasi90, PASI100 = prob_pasi100) |>
  pivot_longer(PASI50:PASI100, names_to = "endpoint", values_to = "p") |>
  mutate(endpoint = factor(endpoint, c("PASI50", "PASI75", "PASI90", "PASI100"))) |>
  ggplot(aes(dose, 100 * p, colour = endpoint)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~arm, scales = "free_x") +
  labs(x = "Dose per administration (mg)", y = "Week 12 responders (%)", colour = NULL) +
  coord_cartesian(ylim = c(0, 100)) +
  theme_bw()
```

## Figure 4: placebo-adjusted treatment effect

```{r fig4, fig.cap = "Replicates Figure 4 of Checchio 2017: Week-12 placebo-adjusted PASI75 and PASI90 treatment effect by drug at the clinical dose, typical 90 kg arm."}
placeboRow <- rxode2::rxSolve(uiLand, makeArmData(arms |> filter(arm == "Placebo"), c(0, 12)),
                              returnType = "data.frame") |>
  filter(time == 12)

solLand |>
  transmute(arm,
            PASI75 = 100 * (prob_pasi75 - placeboRow$prob_pasi75),
            PASI90 = 100 * (prob_pasi90 - placeboRow$prob_pasi90)) |>
  pivot_longer(PASI75:PASI90, names_to = "endpoint", values_to = "effect") |>
  mutate(arm = stats::reorder(factor(arm), effect, FUN = max)) |>
  ggplot(aes(effect, arm, colour = endpoint)) +
  geom_point(size = 2) +
  labs(x = "Placebo-adjusted responders at Week 12 (percentage points)",
       y = NULL, colour = NULL) +
  theme_bw()
```

The source's own headline comparison is that ixekizumab has the highest
placebo-adjusted response, 88.2% - 5.5% = 82.7 percentage points for PASI75
and 70.0 for PASI90. The model gives
`r sprintf("%.1f", 100 * (solLand$prob_pasi75[solLand$arm == "Ixekizumab"] - placeboRow$prob_pasi75))`
and
`r sprintf("%.1f", 100 * (solLand$prob_pasi90[solLand$arm == "Ixekizumab"] - placeboRow$prob_pasi90))`.

```{r fig4-gate}
ixe <- solLand |> filter(arm == "Ixekizumab")
stopifnot(
  abs(100 * (ixe$prob_pasi75 - placeboRow$prob_pasi75) - 82.7) < 1.5,
  abs(100 * (ixe$prob_pasi90 - placeboRow$prob_pasi90) - 70.0) < 1.5,
  # Ixekizumab must be the top agent on both endpoints, as the source reports.
  which.max(solLand$prob_pasi75) == which(solLand$arm == "Ixekizumab"),
  which.max(solLand$prob_pasi90) == which(solLand$arm == "Ixekizumab")
)
```

# Figure 5: external validation of the PASI90 time course

The source's most demanding check bridges the two models: the longitudinal
model, fitted to PASI75 only, is used to predict the **PASI90 time course** by
importing two scaling factors from the landmark model -- additively on the
placebo term and multiplicatively on the drug term. The packaged longitudinal
model carries those two imported constants as `fixed()` parameters and emits
`prob_pasi90` as a secondary output.

```{r fig5, fig.cap = "Replicates Figure 5 of Checchio 2017: PASI90 time course predicted by the longitudinal PASI75 model with the landmark PASI75-to-PASI90 scaling factors applied."}
solLong |>
  filter(time <= 24) |>
  select(arm, time, PASI75 = prob_pasi75, PASI90 = prob_pasi90) |>
  pivot_longer(PASI75:PASI90, names_to = "endpoint", values_to = "p") |>
  filter(arm %in% c("Ixekizumab", "Brodalumab", "Adalimumab", "Ustekinumab",
                    "Etanercept 50", "Methotrexate")) |>
  ggplot(aes(time, 100 * p, colour = arm, linetype = endpoint)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time (weeks)", y = "Responders (%)", colour = NULL, linetype = NULL) +
  coord_cartesian(ylim = c(0, 100)) +
  theme_bw()
```

The bridge is checked against the landmark model rather than eyeballed: the two
models were fitted independently, on different datasets, in different software,
so their Week-12 PASI90 predictions agreeing is a real cross-validation. The
source makes the same claim in prose ("There was close agreement between the
estimates from the longitudinal model and the landmark model"), and the
assertion below is what turns that claim into a test.

```{r fig5-gate}
bridge <- solLong |>
  filter(time == 12) |>
  transmute(arm, longP90 = 100 * prob_pasi90) |>
  inner_join(cmpLand |> transmute(arm, landP90 = p90), by = "arm") |>
  mutate(diff = longP90 - landP90)

bridge |>
  transmute(Arm = arm, `Longitudinal PASI90 (%)` = round(longP90, 1),
            `Landmark PASI90 (%)` = round(landP90, 1),
            `Difference (pp)` = round(diff, 1)) |>
  knitr::kable(caption = "Week-12 PASI90 from the two independently fitted models.")

stopifnot(
  # PASI90 must sit below PASI75 at every time in every arm.
  all(solLong$prob_pasi90 <= solLong$prob_pasi75),
  # Nobody is a PASI90 responder at randomisation.
  all(solLong$prob_pasi90[solLong$time == 0] < 1e-3),
  # The two independently fitted models must agree on the central tendency.
  # Gated on the median and a robust quantile, not the extreme: the two models
  # used different study sets (57 vs 71 trials) and different software, so
  # per-drug scatter is expected and is not a transcription signal.
  abs(stats::median(bridge$diff)) < 5,
  stats::quantile(abs(bridge$diff), 0.9) < 15
)
cat(sprintf("Longitudinal vs landmark Week-12 PASI90: median difference %.1f pp, 90th percentile |difference| %.1f pp\n",
            stats::median(bridge$diff), stats::quantile(abs(bridge$diff), 0.9)))
```

# Between-study variability

The longitudinal model is the one of the pair that carries usable random
effects. Simulating study arms rather than patients shows the spread a trial
designer should expect around the typical trajectory. The cohort is 150 arms,
well inside the 200-per-arm cap.

```{r vpc, fig.cap = "Simulated between-study spread of the ustekinumab 45 mg Q12W PASI75 time course (150 study arms of 100 patients each). The heavy line is the typical-value trajectory."}
rxode2::rxSetSeed(1234)
nArms <- 150
vpcArms <- arms |> filter(arm == "Ustekinumab") |> slice(rep(1, nArms))
vpcDat <- makeArmData(vpcArms, seq(0, 24, by = 1), nArm = 100)
vpcSol <- rxode2::rxSolve(uiLong, vpcDat, returnType = "data.frame")

vpcQ <- vpcSol |>
  group_by(time) |>
  summarise(lo = stats::quantile(prob_pasi75, 0.05),
            md = stats::median(prob_pasi75),
            hi = stats::quantile(prob_pasi75, 0.95), .groups = "drop")
typical <- solLong |> filter(arm == "Ustekinumab", time <= 24, time %% 1 == 0)

ggplot(vpcQ, aes(time)) +
  geom_ribbon(aes(ymin = 100 * lo, ymax = 100 * hi), fill = "grey80") +
  geom_line(aes(y = 100 * md), linewidth = 0.5, linetype = "dashed") +
  geom_line(data = typical, aes(time, 100 * prob_pasi75), linewidth = 1) +
  labs(x = "Time (weeks)", y = "PASI75 responders (%)") +
  coord_cartesian(ylim = c(0, 100)) +
  theme_bw()

stopifnot(
  # The simulated arm distribution must bracket the typical trajectory: with
  # random effects centred at zero, the median arm should track it closely.
  # Gated on the central tendency, not on any simulated extreme, because the
  # draw is not reproducible across rxode2 versions.
  abs(100 * (vpcQ$md[vpcQ$time == 12] - typical$prob_pasi75[typical$time == 12])) < 5,
  # The spread must be non-degenerate and physically plausible.
  vpcQ$hi[vpcQ$time == 12] - vpcQ$lo[vpcQ$time == 12] > 0.05
)
```

# Assumptions and deviations

## Equation recovery

Every display equation is rasterised in the published PDF. Plain-text
extraction of the PDF loses all thirteen and yields no mathematics at all. All
thirteen equations were recovered with
`pdftotext -layout`. Wherever the recovered text was ambiguous, the reading was
settled by whether it reproduces the source's own published predictions, not by
plausibility -- the Table 1 and Table 2 comparisons above are that test.

## The Hill coefficient is absent from the longitudinal model

Equation 4 carries an exponent and Methods says "The Hill coefficient was also
tested in the model", but no value appears anywhere in Supplementary Tables
S1.1 or S1.2. It was therefore tested and not retained, and this model uses
gamma = 1. That reading is confirmed numerically: every Week-4, Week-12 and
ET50 value in Table 1 is reproduced. The landmark model *does* estimate a Hill
coefficient, so the absence is a real difference between the two analyses
rather than a reporting gap.

## Dose is per administration

Neither paper table states the dose metric. It is settled arithmetically, by
pairs of arms that share a single ED50 and differ only in dose: etanercept 25
and 50 mg twice weekly, and tofacitinib 5 and 10 mg twice daily. Only the
per-administration reading reproduces both members of both pairs. Apremilast
is the clearest single case: 30 mg reproduces its published values and 60 mg
does not. Infliximab is mg/kg, per the Supplementary Table S1.1 footnote.

## Certolizumab has no onset rate

Supplementary Table S1.1 reports no `kdrug` for certolizumab, "Not estimated as
only one study for certolizumab was published". The model uses the
kdrug-to-infinity limit of Equation 4 -- zero before time zero, full effect
after. That is not a choice among plausible options: Table 1's certolizumab
Week-4 rate of 38.8% requires a drug term of 4.102 logit units, which already
exceeds that term's own ceiling of 4.097, so no finite onset rate can produce
it. The limit reproduces all four published certolizumab numbers.

## Methotrexate, alefacept and the traditional oral agents are steps

Where the source could not estimate an ED50 it used a single-step offset. In
the longitudinal model that applies to methotrexate and alefacept; in the
landmark model to methotrexate, ciclosporin and acitretin. For those agents the
model is only valid at the clinical dose in the arm table, and `> 0` is the only
thing read from the dose column. Predicting a different dose of those agents
from these models is not supported.

## The alefacept route

Supplementary Table 2 carries an IV offset on the alefacept ED50 with a non-IV
(intramuscular) reference. The source's own Table 2 prediction for "Alefacept
10 mg QW" is reproduced by the **IV** branch (24.1% against a published 22.7%),
not the IM branch (11.8%). `ROUTE_IV = 1` is therefore required to reproduce
any published alefacept number, and the arm table above sets it. The source
does not state which route its typical-patient simulation assumed; the
arithmetic does.

## Two values are back-solved, not printed

`e0_trial` and `led50_briakinumab` are not in the paper or its supplement and
are recovered by inverting the source's own Table 2 predictions through its own
Equations 5-8, as shown above and re-derived at render time. Neither is a value
imported from outside the paper. The evidence that the inversion is sound is
that `e0_trial` is over-determined 36-to-1 with small unstructured residuals,
and that it independently reproduces the typical placebo rates implied by
Table 2's separate "difference from placebo" columns.

## Random-effect and residual scale in the longitudinal model

Supplementary Table S1.2 labels its variance rows "omega", "sigma" and
"sigma corr" while the Methods text names the variances "omega squared" and
"sigma squared", so the printed numbers could be read as variances or as
standard deviations. They are read here as **variances**, for two reasons. The
table is headed "Common Parameter Estimates from NONMEM Output", and a NONMEM
`$OMEGA` estimate is a variance. And the SD reading is physically too tight:
it would confine the Week-12 placebo PASI75 to 4.0-5.5% across a one-SD band,
whereas published psoriasis placebo arms span roughly 1-8% and the source's own
landmark analysis measures an I-squared of 82% for between-study placebo
heterogeneity. The variance reading gives 2.3-9.4%.

This choice changes **no typical-value prediction** -- every comparison in this
vignette is unaffected -- only the simulated between-study spread in the section
above. For the two residual terms it is nearly immaterial even there: the total
multiplier on the binomial standard error is 1.28 under the variance reading and
1.21 under the SD reading, and only the split between the two terms differs.

## The landmark model has no variability

Equation 5 carries an endpoint-specific random effect and the Methods describe
within-study and between-arm correlation terms, but Supplementary Table 2
tabulates only fixed effects: no variance, no correlation matrix, and no
residual. None is encoded, because inventing one would be unauditable. The
landmark model therefore produces typical-value arm predictions only, and its
`ini()` carries a fixed placeholder residual purely so the likelihood machinery
accepts it for forward simulation. Use the longitudinal model when
between-study spread is needed.

The practical consequence is visible in the Table 2 comparison: the source's
published numbers are means over 1,000 simulations *including* that random
effect, so the two arms with the largest residuals differ from the typical-value
prediction in the direction Jensen's inequality requires.

## ET90 runs short for the slower drugs

Reproduced ET50 matches every published value to within 0.3 weeks. ET90 runs
about one week short for the slowest drugs, most likely because the source
evaluated the maximal treatment effect at a finite horizon rather than at
infinity. ET50 is the metric the source's Discussion uses; the ET90 deviation is
recorded here rather than tuned away.

## Residual weighting is applied inside the model

Most of this package's MBMA extractions store the residual SD unweighted and
leave the `1/sqrt(N)` arm weighting to downstream code. This one carries `N_ARM`
as a covariate column and applies the binomial standard error
`W = sqrt(Pr*(1-Pr)/N_ARM)` inside `model()`, because the arm-level correlated
term of Equation 13 is a *random effect* whose SD varies from arm to arm, and a
per-arm random-effect SD cannot be reproduced by rescaling the output of a
single `rxSolve()`.

## Scope

Both models predict **study-arm mean responder rates**, not individual-patient
outcomes. The source is explicit that the body weight effect is an
aggregate-level association whose magnitude is probably attenuated, and that its
attribution to the placebo component "should not be interpreted in a mechanistic
context". Neither model has a PK layer, so neither can be driven by exposure.

## Bootstrap versus final estimates

The Results text quotes the body weight effects as -0.193 and -0.867, which are
**medians of a 1,000-replicate nonparametric bootstrap**. Supplementary Table
S1.2 gives the NONMEM final estimates, -0.198 and -0.834, and those are what the
model uses. Both pairs exclude zero and agree in sign and magnitude.
