---
title: "Vancomycin (Chen 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vancomycin (Chen 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Chen_2023_vancomycin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Chen_2023_vancomycin"))$description`
* Article: <https://doi.org/10.1093/jac/dkad202>

# Population

Chen 2023 is a retrospective single-centre therapeutic-drug-monitoring study of
673 children treated with intravenous vancomycin at Xinhua Hospital, Shanghai,
between June 2013 and June 2022, contributing 1547 serum concentrations (779
troughs and 768 peaks). Postnatal age spanned 0.01 to 17.23 years (median 1.81,
IQR 0.51-5.83) and body weight 1.4 to 69.0 kg (median 11.5, IQR 6.5-20.0);
59.0% were male. 82.0% were ICU inpatients and 10.1% had undergone
cardiothoracic surgery and were receiving vasoactive agents. Renal function
spanned a wide range (eGFR by the paper's Equation 11: median 109
mL/min/1.73 m^2, IQR 87-141, range 12-379), with 1.0% below 30 and 38.5% above
120. The leading indications were pneumonia (40.6%), primary bloodstream
infection (40.0%) and CNS infection (26.0%). Patients on renal replacement
therapy or ECMO were excluded. Baseline demographics are Chen 2023 Table 1.

A separate 53-subject external-validation cohort was recruited at Shanghai
Children's Medical Center; it contained no cardiothoracic-surgery patients and
only a single peak concentration, and was not used for estimation.

The same information is available programmatically via
`readModelDb("Chen_2023_vancomycin")()$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Chen_2023_vancomycin.R`. The table below
collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `CL = TV(CL) * (WT/70)^0.75 * MF * (eGFR/109)^theta_eGFR * theta_1^CTS` | n/a | Table 3, "The final model", first equation |
| `V  = TV(V) * (WT/70) * (36.2/ALB)^theta_ALB` | n/a | Table 3, "The final model", second equation |
| `MF = 1 / (1 + (PMA/TM50)^Hill)` | n/a | Equation 6 (maturation model III) |
| `eGFR = 40.7 * (height(m)/SCr)^0.64 * (30/BUN)^0.202` | n/a | Equation 11; restated in Table 3, "The final model", third equation |
| eGFR cap at 120 mL/min/1.73 m^2 | 120 | Table 2 Model D ("eGFR 1 (Equation 11) with cap 120"); Results, "PPK modelling" |
| Allometric exponent on CL | 0.75 (fixed) | Equations 1 and 5, printed without uncertainty |
| Allometric exponent on V | 1 (fixed) | Equation 2; Methods after Equation 10: "allometric exponents for volume of distribution (V) were fixed to 1" |
| `lcl` = TV(CL) | 7.75 L/h | Table 3 (RSE 2.3%; bootstrap 95% CI 7.43-8.13) |
| `lvc` = TV(V) | 36.2 L | Table 3 (RSE 1.7%; bootstrap 95% CI 35.0-37.4) |
| `pma_tm50` = TM50 | 37.0 weeks | Table 3 (RSE 4.6%; bootstrap 95% CI 33.5-40.2) |
| `pma_hill` = Hill | -1.63 | Table 3 (RSE 13%; bootstrap 95% CI -2.11 to -1.24) |
| `e_crcl_cl` = theta_eGFR | 1.01 | Table 3 (RSE 3.4%; bootstrap 95% CI 0.94-1.08) |
| `e_cts_cl` = theta_1 | 0.761 | Table 3 (RSE 3.6%; bootstrap 95% CI 0.709-0.817) |
| `e_alb_vc` = theta_ALB | 0.279 | Table 3 (RSE 26.7%; bootstrap 95% CI 0.135-0.423) |
| eGFR reference | 109 mL/min/1.73 m^2 | Table 3 final-model equation; equals the Table 1 model-group median for Equation 11 |
| ALB reference | 36.2 g/L | Table 3 final-model equation (Table 1 median 36.3 g/L) |
| IIV on CL | 27.5 %CV | Table 3 (RSE 6.0%) |
| IIV on V | 27.2 %CV | Table 3 (RSE 16.2%) |
| Cov(CL, V) | 0.0536 | Table 3 (RSE 10.8%) |
| `propSd` (exponential arm) | 17.7 %CV | Table 3 (RSE 10.6%); see Assumptions |
| `addSd` | 0.154 mg/L | Table 3 (RSE 44.5%) |

```{r load-model}
mod <- readModelDb("Chen_2023_vancomycin")
# Typical-value form: zero the between-subject variability but KEEP the
# residual-error parameters, which the error model still needs to resolve.
mod_tv <- rxode2::zeroRe(mod, which = "omega")
```

```{r helpers}
# One dosing-regimen event table for one subject, expressed as a plain data
# frame. `cmt` is set explicitly to the ODE STATE name (`central`) on every
# row -- never to the algebraic observable `Cc`, which would auto-inject a
# compartment slot and renumber the model.
# `days` defaults to 10 so that even the slowest-clearing subject the cohort can
# draw is at steady state by the final interval. The cohort's floor eGFR of 12
# mL/min/1.73 m^2 in a neonate gives a terminal half-life near 27 h, so 3 days
# of dosing leaves such a subject several percent short of steady state and
# breaks the AUC24 = dose/CL identity below. Only dosing records are added --
# observations stay on the final interval -- so the extra days are nearly free.
make_events <- function(id, WT, PAGE, CRCL, ALB, CTS, mg_kg_day, tau,
                        days = 10, obs_by = 0.1, tinf = 1) {
  amt <- WT * mg_kg_day / (24 / tau)
  last_dose <- days * 24 - tau
  dose <- data.frame(
    id = id, time = seq(0, last_dose, by = tau), amt = amt, evid = 1,
    rate = amt / tinf
  )
  obs <- data.frame(
    id = id, time = seq(last_dose, last_dose + tau, by = obs_by),
    amt = NA_real_, evid = 0, rate = NA_real_
  )
  out <- rbind(dose, obs)
  out$cmt <- "central"
  out$WT <- WT; out$PAGE <- PAGE; out$CRCL <- CRCL; out$ALB <- ALB
  out$SURG_CARD_VASOACTIVE <- CTS
  out$tau <- tau; out$mg_kg_day <- mg_kg_day; out$last_dose <- last_dose
  out[order(out$time, out$evid), ]
}

# Individual clearance and volume for one covariate combination, read straight
# off the typical-value solve.
typical_pk <- function(WT, PAGE, CRCL, ALB, CTS) {
  ev <- make_events(1L, WT, PAGE, CRCL, ALB, CTS,
                    mg_kg_day = 40, tau = 24, days = 1, obs_by = 1)
  s <- rxode2::rxSolve(mod_tv, ev, returnType = "data.frame")
  c(cl = s$cl[1], vc = s$vc[1], mf = s$maturation_cl[1],
    crcl_capped = s$crcl_capped[1])
}
```

# Structural identities

These checks are deterministic: they use the typical-value model with no
between-subject variability, so each is an exact algebraic identity of the
published equations rather than a cohort statistic. They are gated tightly.

```{r identities}
tol <- 1e-8

# 1. Maturation is exactly 50% at PMA = TM50 = 37.0 weeks (Chen 2023 Methods:
#    "TM50 is the PMA at which CL maturation reaches 50% of that of adults").
mf_at_tm50 <- typical_pk(10, 37.0, 109, 36.2, 0)[["mf"]]

# 2. The eGFR ceiling: any eGFR above 120 must behave exactly like 120
#    (Chen 2023 Table 2 Model D).
cl_120 <- typical_pk(10, 200, 120, 36.2, 0)[["cl"]]
cl_200 <- typical_pk(10, 200, 200, 36.2, 0)[["cl"]]
cl_379 <- typical_pk(10, 200, 379, 36.2, 0)[["cl"]]  # cohort maximum eGFR

# 3. Cardiothoracic surgery with vasoactive support multiplies CL by exactly
#    theta_1 = 0.761, i.e. the "roughly 24% reduction" of the Discussion.
cts_ratio <- typical_pk(10, 200, 109, 36.2, 1)[["cl"]] /
  typical_pk(10, 200, 109, 36.2, 0)[["cl"]]

# 4. At the reference albumin the volume term collapses to pure allometry:
#    V = 36.2 * (WT/70).
vc_ref <- typical_pk(10, 200, 109, 36.2, 0)[["vc"]]

identities <- tibble::tibble(
  Check = c(
    "Maturation factor at PMA = TM50 (37.0 weeks)",
    "CL at eGFR 200 equals CL at eGFR 120 (cap)",
    "CL at eGFR 379 equals CL at eGFR 120 (cap)",
    "CL(CTS) / CL(non-CTS)",
    "V at ALB = 36.2 g/L equals 36.2 * (WT/70)"
  ),
  Expected = c(0.5, cl_120, cl_120, 0.761, 36.2 * 10 / 70),
  Achieved = c(mf_at_tm50, cl_200, cl_379, cts_ratio, vc_ref)
) |>
  mutate(`Abs. difference` = abs(Achieved - Expected))

knitr::kable(identities, digits = 10,
             caption = "Deterministic structural identities of the Chen 2023 equations.")

stopifnot(all(identities$`Abs. difference` < tol))
```

# The covariate model

```{r covariate-curves, fig.height = 3.4}
mat_curve <- tibble::tibble(PMA = seq(24, 600, by = 2)) |>
  mutate(MF = vapply(PMA, function(p) typical_pk(10, p, 109, 36.2, 0)[["mf"]], numeric(1)))

p_mat <- ggplot(mat_curve, aes(PMA, MF)) +
  geom_line() +
  geom_hline(yintercept = 0.5, linetype = "dotted") +
  geom_vline(xintercept = 37, linetype = "dotted") +
  scale_x_log10() +
  labs(x = "Postmenstrual age (weeks, log scale)", y = "Maturation factor",
       title = "Sigmoid PMA maturation on CL",
       subtitle = "Chen 2023 Equation 6; TM50 = 37.0 weeks, Hill = -1.63")

egfr_curve <- tibble::tibble(CRCL = seq(10, 380, by = 2)) |>
  mutate(CL = vapply(CRCL, function(g) typical_pk(10, 200, g, 36.2, 0)[["cl"]], numeric(1)))

p_egfr <- ggplot(egfr_curve, aes(CRCL, CL)) +
  geom_line() +
  geom_vline(xintercept = 120, linetype = "dotted") +
  labs(x = expression("eGFR (mL/min/1.73 m"^2*")"), y = "CL (L/h), 10 kg mature child",
       title = "Renal-function effect on CL, capped at 120",
       subtitle = "Chen 2023 Table 2 Model D")

print(p_mat)
print(p_egfr)
```

## Replicating Figure 1

Chen 2023 Figure 1 plots individual clearance in L/h/kg against eGFR, with
separate smooths for cardiothoracic-surgery and non-cardiothoracic-surgery
patients and a vertical marker at eGFR 120. The model reproduces both features:
the CTS curve sits uniformly below the non-CTS curve, and both flatten
completely above 120 because of the cap.

```{r figure-1, fig.height = 4}
fig1 <- tidyr::crossing(CRCL = seq(10, 380, by = 5), CTS = c(0, 1)) |>
  rowwise() |>
  mutate(cl_per_kg = typical_pk(11.5, 134, CRCL, 36.3, CTS)[["cl"]] / 11.5) |>
  ungroup() |>
  mutate(Group = ifelse(CTS == 1, "CTS patients", "Non-CTS patients"))

ggplot(fig1, aes(CRCL, cl_per_kg, colour = Group, linetype = Group)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = 120, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c("CTS patients" = "darkred",
                                 "Non-CTS patients" = "steelblue")) +
  labs(x = expression("eGFR (mL/min/1.73 m"^2*")"), y = "Vancomycin CL (L/h/kg)",
       title = "Figure 1 - vancomycin CL versus eGFR",
       caption = paste("Replicates Figure 1 of Chen 2023 (typical 11.5 kg,",
                       "PMA 134-week child). The dashed line marks eGFR 120.")) +
  theme(legend.position = "bottom")

# The cap must make both curves exactly flat above 120, and the CTS curve must
# sit below the non-CTS curve everywhere. Both are algebraic consequences of
# the equations, so they are gated exactly.
flat <- fig1 |> filter(CRCL >= 120) |> group_by(Group) |>
  summarise(spread = diff(range(cl_per_kg)), .groups = "drop")
stopifnot(all(flat$spread < 1e-10))
wide <- fig1 |> select(CRCL, Group, cl_per_kg) |>
  tidyr::pivot_wider(names_from = Group, values_from = cl_per_kg)
stopifnot(all(wide$`CTS patients` < wide$`Non-CTS patients`))
```

# Replicating Table 5: the published dosing recommendations

This is the strongest available check on the transcription. Chen 2023 Table 5
gives an optimal daily dose for each combination of age band, cardiothoracic-
surgery status and eGFR band, chosen so that most patients reach an AUC24/MIC
of 400-650 at steady state with an assumed MIC of 1 mg/L. The paper computes
AUC24 as total daily dose divided by CL (Methods, "Simulations"), so a
correctly transcribed clearance model must place a typical patient given each
of those 32 recommended doses inside the 400-650 window.

The representative body weight and postmenstrual age for each age band are
chosen here, not taken from the paper, so the agreement below is not
circular - it tests the clearance equation across a 13-fold weight range, a
20-fold PMA range, both surgery states and all four renal strata at once.

```{r table5}
reps <- tibble::tribble(
  ~age_group,            ~WT,  ~PAGE,
  "0 to <1 month",        3.5,   42,
  "1 month to <2 years",  8.0,   70,
  "2 to <12 years",      18.0,  320,
  "12 to <18 years",     45.0,  750
)

# Chen 2023 Table 5, Dose (mg/kg/day) column. Where the paper gives a range
# (e.g. "25-30") the midpoint is used.
tab5 <- tibble::tribble(
  ~age_group,             ~CTS, ~egfr_band, ~dose,
  "0 to <1 month",           0, "30-60",     27.5,
  "0 to <1 month",           0, "60-90",     47.5,
  "0 to <1 month",           0, "90-120",    60.0,
  "0 to <1 month",           0, ">120",      75.0,
  "0 to <1 month",           1, "30-60",     20.0,
  "0 to <1 month",           1, "60-90",     35.0,
  "0 to <1 month",           1, "90-120",    47.5,
  "0 to <1 month",           1, ">120",      55.0,
  "1 month to <2 years",     0, "30-60",     35.0,
  "1 month to <2 years",     0, "60-90",     50.0,
  "1 month to <2 years",     0, "90-120",    70.0,
  "1 month to <2 years",     0, ">120",      80.0,
  "1 month to <2 years",     1, "30-60",     25.0,
  "1 month to <2 years",     1, "60-90",     40.0,
  "1 month to <2 years",     1, "90-120",    55.0,
  "1 month to <2 years",     1, ">120",      60.0,
  "2 to <12 years",          0, "30-60",     32.5,
  "2 to <12 years",          0, "60-90",     55.0,
  "2 to <12 years",          0, "90-120",    70.0,
  "2 to <12 years",          0, ">120",      80.0,
  "2 to <12 years",          1, "30-60",     25.0,
  "2 to <12 years",          1, "60-90",     40.0,
  "2 to <12 years",          1, "90-120",    52.5,
  "2 to <12 years",          1, ">120",      62.5,
  "12 to <18 years",         0, "30-60",     25.0,
  "12 to <18 years",         0, "60-90",     42.5,
  "12 to <18 years",         0, "90-120",    65.0,
  "12 to <18 years",         0, ">120",      70.0,
  "12 to <18 years",         1, "30-60",     20.0,
  "12 to <18 years",         1, "60-90",     32.5,
  "12 to <18 years",         1, "90-120",    47.5,
  "12 to <18 years",         1, ">120",      52.5
)

egfr_mid <- c("30-60" = 45, "60-90" = 75, "90-120" = 105, ">120" = 140)

tab5 <- tab5 |>
  left_join(reps, by = "age_group") |>
  mutate(CRCL = unname(egfr_mid[egfr_band])) |>
  rowwise() |>
  mutate(cl = typical_pk(WT, PAGE, CRCL, 36.3, CTS)[["cl"]]) |>
  ungroup() |>
  mutate(
    AUC24 = WT * dose / cl,
    in_target = AUC24 >= 400 & AUC24 <= 650
  )

tab5 |>
  mutate(Surgery = ifelse(CTS == 1, "CTS", "Non-CTS")) |>
  select(`Age band` = age_group, Surgery, `eGFR band` = egfr_band,
         `Dose (mg/kg/day)` = dose, `CL (L/h)` = cl,
         `AUC24 (mg*h/L)` = AUC24, `In 400-650` = in_target) |>
  knitr::kable(
    digits = c(0, 0, 0, 1, 3, 0, 0),
    caption = paste("Replicates Chen 2023 Table 5. Steady-state AUC24 for a",
                    "representative patient in each band, given the paper's own",
                    "recommended dose. MIC is assumed to be 1 mg/L, so AUC24",
                    "equals AUC24/MIC.")
  )

# Deterministic given the representative covariates above, so gated on the
# paper's own target window rather than on a tolerance. Realised range is
# reported below so a future reader can see the headroom.
stopifnot(all(tab5$in_target))
```

All `r nrow(tab5)` of the paper's recommended doses land inside the published
400-650 target, with realised AUC24 spanning
`r sprintf("%.0f", min(tab5$AUC24))` to `r sprintf("%.0f", max(tab5$AUC24))`
mg*h/L - clustered near the centre of the window rather than scraping its
edges. A mis-transcribed clearance term, allometric reference, maturation
constant or eGFR exponent would move these by tens of percent and break the
check immediately.

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the Chen 2023 Table 1 marginal distributions: the postnatal-age band
proportions, a weight-for-age relationship with log-normal scatter (so weight
and postmenstrual age stay physiologically coupled), and log-normal eGFR and
normal albumin matched to the published medians and interquartile ranges. The
cardiothoracic-surgery prevalence and the daily-dose distribution follow
Table 1 as well.

```{r cohort}
# set.seed() seeds R's RNG for the covariate draw below. It does NOT seed
# rxode2's simulation RNG, whose streams are partitioned per solver thread --
# so the residual-error and IIV draws differ between a 16-thread workstation
# and a 2-core CI runner. Every assertion downstream is written to hold for any
# cohort the model can produce.
set.seed(20230628)
n_subj <- 200

# Table 1 PNA bands and their proportions.
band_p  <- c(6.1, 31.4, 14.1, 43.5, 4.9) / 100
band_lo <- c(0.01, 1 / 12, 1, 2, 12)
band_hi <- c(1 / 12, 1, 2, 12, 17.83)
band <- sample(seq_along(band_p), n_subj, replace = TRUE, prob = band_p)
pna_years <- band_lo[band] + runif(n_subj) * (band_hi[band] - band_lo[band])

# Weight-for-age: a smooth infant-through-adolescent median with log-normal
# scatter, then clipped to the Table 1 observed range.
wt_median <- 3.3 + 6.2 * pna_years / (0.35 + pna_years) + 2.3 * pna_years
iqr_z <- 2 * stats::qnorm(0.75)

cohort <- tibble::tibble(
  id  = seq_len(n_subj),
  PNA = pna_years,
  WT  = pmin(pmax(wt_median * exp(stats::rnorm(n_subj, 0, 0.22)), 1.4), 69.0),
  # Chen 2023 Methods: PMA = GA + PNA, with GA imputed at 40 weeks above 1 year.
  PAGE = pna_years * 52.18 + 40,
  CRCL = pmin(pmax(exp(stats::rnorm(n_subj, log(109),
                                    (log(141) - log(87)) / iqr_z)), 12), 379),
  ALB  = pmin(pmax(stats::rnorm(n_subj, 36.3, (40.5 - 31.5) / iqr_z), 18.6), 57.9),
  SURG_CARD_VASOACTIVE = stats::rbinom(n_subj, 1, 0.101),
  tau = sample(c(6, 8, 12), n_subj, replace = TRUE),
  mg_kg_day = pmin(pmax(exp(stats::rnorm(n_subj, log(40), 0.22)), 19.7), 70.0)
)

tibble::tibble(
  Characteristic = c("WT (kg)", "PMA (weeks)", "eGFR (mL/min/1.73 m^2)",
                     "ALB (g/L)", "Daily dose (mg/kg)"),
  Simulated = c(
    sprintf("%.1f [%.1f-%.1f]", median(cohort$WT),
            quantile(cohort$WT, .25), quantile(cohort$WT, .75)),
    sprintf("%.0f [%.0f-%.0f]", median(cohort$PAGE),
            quantile(cohort$PAGE, .25), quantile(cohort$PAGE, .75)),
    sprintf("%.0f [%.0f-%.0f]", median(cohort$CRCL),
            quantile(cohort$CRCL, .25), quantile(cohort$CRCL, .75)),
    sprintf("%.1f [%.1f-%.1f]", median(cohort$ALB),
            quantile(cohort$ALB, .25), quantile(cohort$ALB, .75)),
    sprintf("%.1f [%.1f-%.1f]", median(cohort$mg_kg_day),
            quantile(cohort$mg_kg_day, .25), quantile(cohort$mg_kg_day, .75))
  ),
  `Chen 2023 Table 1` = c("11.5 [6.5-20.0]", "134 [66-344]", "109 [87-141]",
                          "36.3 [31.5-40.5]", "40.0 [39.2-57.8]")
) |>
  knitr::kable(caption = "Virtual cohort versus Chen 2023 Table 1 (median [IQR]).")
```

```{r cohort-events}
build_events <- function(df, dose_col, obs_by = 0.1) {
  do.call(rbind, lapply(seq_len(nrow(df)), function(i) {
    make_events(df$id[i], df$WT[i], df$PAGE[i], df$CRCL[i], df$ALB[i],
                df$SURG_CARD_VASOACTIVE[i], df[[dose_col]][i], df$tau[i],
                obs_by = obs_by)
  }))
}

events <- build_events(cohort, "mg_kg_day")

# IDs are already disjoint (one row-block per subject), but assert it anyway:
# duplicate (id, time, evid) triples silently merge subjects in rxSolve.
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

Ten days of dosing brings every subject the cohort can draw to steady state -
the typical terminal half-life is a few hours, and the slowest subject at the
cohort's floor eGFR is near 27 h - and the final dosing interval is observed on
a 0.1 h grid.

```{r simulate}
rxode2::rxSetSeed(20230628)
sim <- rxode2::rxSolve(
  mod, events = events,
  keep = c("tau", "mg_kg_day", "last_dose", "WT", "CRCL",
           "SURG_CARD_VASOACTIVE"),
  returnType = "data.frame"
) |>
  filter(!is.na(Cc)) |>
  mutate(tad = time - last_dose)

stopifnot(nrow(sim) > 0, all(sim$Cc >= 0))
```

```{r vpc, fig.height = 4}
sim |>
  mutate(`Dosing interval` = paste0("q", tau, "h")) |>
  group_by(`Dosing interval`, tad) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = quantile(Cc, 0.50),
            Q95 = quantile(Cc, 0.95), .groups = "drop") |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~`Dosing interval`, scales = "free_x") +
  labs(x = "Time after dose (h)", y = "Vancomycin concentration (mg/L)",
       title = "Steady-state concentration-time profile by dosing interval",
       subtitle = "Median with 5th-95th percentile band, 200 virtual subjects")
```

# PKNCA validation

NCA is run over the final, steady-state dosing interval, grouped by dosing
interval so per-group results are separable.

```{r pknca}
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  mutate(regimen = paste0("q", tau, "h")) |>
  select(id, time, Cc, regimen, tau, last_dose)

dose_df <- events |>
  filter(evid == 1) |>
  mutate(regimen = paste0("q", tau, "h")) |>
  select(id, time, amt, regimen)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(sim_nca), Cc ~ time | regimen + id,
                             concu = "mg/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | regimen + id)

# One steady-state interval per subject: the last dosing interval.
intervals <- sim_nca |>
  distinct(id, regimen, tau, last_dose) |>
  transmute(start = last_dose, end = last_dose + tau,
            regimen, id,
            cmax = TRUE, tmax = TRUE, cmin = TRUE,
            auclast = TRUE, cav = TRUE) |>
  as.data.frame()

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
nca_tbl <- as.data.frame(nca_res$result)
```

## Internal identity: AUC24 equals daily dose divided by clearance

For a linear one-compartment model at steady state, the area under the curve
across one dosing interval, scaled to 24 h, equals the total daily dose divided
by clearance. That is exactly the relationship Chen 2023 uses to convert a dose
into an AUC24/MIC (Methods, "Simulations"), so it is worth confirming that the
packaged model honours it. Both sides here come from the same simulation, so
the only difference is trapezoidal-integration error on a 0.1 h grid; the check
is gated tightly.

```{r auc-identity}
auc_check <- nca_tbl |>
  filter(PPTESTCD == "auclast") |>
  select(id, regimen, auclast = PPORRES) |>
  left_join(cohort |> select(id, WT, mg_kg_day, tau), by = "id") |>
  left_join(
    sim |> distinct(id, cl_ind = cl),
    by = "id"
  ) |>
  mutate(
    auc24_nca = auclast * 24 / tau,
    auc24_identity = WT * mg_kg_day / cl_ind,
    pct_diff = 100 * (auc24_nca - auc24_identity) / auc24_identity
  )

tibble::tibble(
  Statistic = c("Median |% difference|", "Maximum |% difference|"),
  Value = c(median(abs(auc_check$pct_diff)), max(abs(auc_check$pct_diff)))
) |>
  knitr::kable(digits = 4,
               caption = "AUC24 from PKNCA versus daily dose / CL.")

# Trapezoidal error on a 0.1 h grid only; nothing stochastic enters the
# comparison because both sides use the same drawn parameters, so this is
# gated near the accuracy actually achieved rather than loosely. Realised max
# 0.0143% at both 16 and 2 solver threads; 0.2 keeps an order of magnitude of
# headroom for an unluckier draw while still going red on any real structural
# error (a missed dose, a wrong interval, a window that is not yet at steady
# state all move this by whole percent -- a 3-day run-in put it over 1%).
stopifnot(max(abs(auc_check$pct_diff)) < 0.2)
```

## Comparison against the published concentrations

Chen 2023 Table 1 reports the observed peak and trough concentrations of the
model group. Its footnote c defines these by **sampling window**, not by
extremum: "Peak: between 0 and 2 h after the end of infusion; trough: >2 h
after the end of infusion." A trough sampled anywhere more than 2 h after the
end of a 1 h infusion is therefore not a pre-dose trough, and comparing the
published value against a true pre-dose concentration understates the model by
roughly 40%. The table below shows both readings so the distinction is visible:
the "NCA extremes" rows use PKNCA's Cmax and Cmin over the dosing interval,
while the "Table 1 sampling windows" rows re-sample the same simulation using
the paper's own window definitions.

```{r nca_compare}
# Re-sample each subject once inside each of the paper's two windows.
set.seed(11)
windowed <- sim |>
  group_by(id) |>
  summarise(
    cmax = {
      w <- Cc[tad >= 1 & tad <= 3]
      if (length(w)) sample(w, 1) else NA_real_
    },
    cmin = {
      w <- Cc[tad > 3 & tad <= first(tau)]
      if (length(w)) sample(w, 1) else NA_real_
    },
    .groups = "drop"
  ) |>
  filter(!is.na(cmax), !is.na(cmin))
stopifnot(nrow(windowed) > 100)

simulated_long <- bind_rows(
  nca_tbl |>
    filter(PPTESTCD %in% c("cmax", "cmin")) |>
    transmute(sampling = "NCA extremes over the dosing interval",
              PPTESTCD, PPORRES),
  windowed |>
    tidyr::pivot_longer(c(cmax, cmin), names_to = "PPTESTCD",
                        values_to = "PPORRES") |>
    mutate(sampling = "Chen 2023 Table 1 sampling windows")
)

published <- tibble::tribble(
  ~sampling,                                 ~cmax,  ~cmin,
  "NCA extremes over the dosing interval",   19.91,  7.18,
  "Chen 2023 Table 1 sampling windows",      19.91,  7.18
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = as.data.frame(simulated_long),
  reference     = as.data.frame(published),
  by            = "sampling",
  units         = c(cmax = "mg/L", cmin = "mg/L"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  digits  = 2,
  caption = paste("Simulated versus Chen 2023 Table 1 peak and trough",
                  "concentrations (medians). Reference values are identical",
                  "across both rows - only the simulated sampling scheme",
                  "differs. * differs from reference by >20%.")
)
```

Read against the paper's own sampling windows, the simulated peak and trough
medians track the published values closely; read as NCA extremes they bracket
them, which is the expected direction given that a window sample can only fall
below the interval maximum and above the interval minimum.

```{r compare-gate}
# `cmp`'s "% diff" column is a formatted string carrying the >20% asterisk, so
# the gate recomputes the window-matched differences from the data rather than
# parsing it back out.
ref_peak   <- 19.91  # Chen 2023 Table 1, model-group median peak (mg/L)
ref_trough <- 7.18   # Chen 2023 Table 1, model-group median trough (mg/L)

win_peak   <- 100 * (median(windowed$cmax) - ref_peak)   / ref_peak
win_trough <- 100 * (median(windowed$cmin) - ref_trough) / ref_trough

tibble::tibble(
  Quantity = c("Peak (window 0-2 h post-infusion)",
               "Trough (window >2 h post-infusion)"),
  `Chen 2023 Table 1` = c(ref_peak, ref_trough),
  `Simulated median` = c(median(windowed$cmax), median(windowed$cmin)),
  `% difference` = c(win_peak, win_trough)
) |>
  knitr::kable(digits = 1,
               caption = "Window-matched simulated concentrations versus Chen 2023 Table 1.")

# These ARE cohort statistics, so the bound has to survive any cohort the model
# can draw and any solver-thread count. Two effects sit inside it that no
# amount of correct transcription removes: the virtual cohort matches Table 1
# marginally but not jointly, and the published values come from clinically
# triggered TDM sampling rather than a randomised design. Measured across
# solver-thread counts, which draw different cohorts: peak +16.5% and trough
# +32.4% at 16 threads, peak +11.0% and trough +25.8% at 2 threads. Do not
# tighten 60 towards those: a mis-transcribed CL, V, dose or unit moves a
# concentration by a factor rather than by tens of percent, so 60 still goes
# red on any such error while admitting the cohort noise above.
stopifnot(abs(win_peak) < 60, abs(win_trough) < 60)
```

## Optimal trough range

Chen 2023 Figure 6 concludes that, under the Table 5 dosing regimens, the
trough concentration corresponding to an AUC24/MIC of 400-650 falls inside a
narrower band within 8-15 mg/L. The check below doses each virtual subject at
their band's recommended dose and reports the resulting pre-dose trough
distribution.

```{r trough-range}
band_of <- function(egfr) {
  cut(pmin(egfr, 200), c(-Inf, 60, 90, 120, Inf),
      labels = c("30-60", "60-90", "90-120", ">120"))
}
age_band_of <- function(pna) {
  cut(pna, c(-Inf, 1 / 12, 2, 12, Inf),
      labels = c("0 to <1 month", "1 month to <2 years",
                 "2 to <12 years", "12 to <18 years"))
}

opt <- cohort |>
  mutate(egfr_band = as.character(band_of(CRCL)),
         age_group = as.character(age_band_of(PNA))) |>
  filter(CRCL >= 30) |>
  left_join(tab5 |> select(age_group, CTS, egfr_band, opt_dose = dose),
            by = c("age_group", "egfr_band",
                   "SURG_CARD_VASOACTIVE" = "CTS"))
stopifnot(!anyNA(opt$opt_dose))

opt_events <- build_events(opt, "opt_dose", obs_by = 0.25)

rxode2::rxSetSeed(99)
opt_sim <- rxode2::rxSolve(mod, events = opt_events,
                           keep = c("tau", "mg_kg_day", "last_dose"),
                           returnType = "data.frame") |>
  filter(!is.na(Cc)) |>
  mutate(tad = time - last_dose)

trough <- opt_sim |>
  group_by(id) |>
  filter(abs(tad - first(tau)) < 1e-6) |>
  ungroup()
auc24 <- opt_sim |>
  distinct(id, cl, mg_kg_day) |>
  left_join(cohort |> select(id, WT), by = "id") |>
  mutate(AUC24 = WT * mg_kg_day / cl)

tibble::tibble(
  Quantity = c("Pre-dose trough (mg/L)", "AUC24/MIC (MIC = 1 mg/L)"),
  `Median [IQR]` = c(
    sprintf("%.1f [%.1f-%.1f]", median(trough$Cc),
            quantile(trough$Cc, .25), quantile(trough$Cc, .75)),
    sprintf("%.0f [%.0f-%.0f]", median(auc24$AUC24),
            quantile(auc24$AUC24, .25), quantile(auc24$AUC24, .75))
  ),
  `Chen 2023` = c("8-15 (Figure 6, optimal range)",
                  "400-650 (target, Table 5)")
) |>
  knitr::kable(caption = paste("Virtual cohort dosed at its Chen 2023 Table 5",
                               "recommended dose."))

# The cohort median AUC24 must land in the paper's own target window. This is a
# cohort statistic, but the window is 400-650 and the realised median is 524
# at 16 solver threads and 500 at 2, both near its centre, so there is real
# headroom on both sides.
stopifnot(median(auc24$AUC24) > 400, median(auc24$AUC24) < 650)
```

The cohort median AUC24/MIC of `r sprintf("%.0f", median(auc24$AUC24))` sits
near the centre of the paper's 400-650 target, which is the substantive claim
of Table 5. The pre-dose trough distribution, median
`r sprintf("%.1f", median(trough$Cc))` mg/L, sits a little below the 8-15 mg/L
band of Figure 6. That is expected rather than contradictory: Figure 6 is built
on the paper's own resampled virtual population, whose joint covariate
structure this vignette reproduces only marginally (see Assumptions), and the
trough is far more sensitive than the AUC to the volume term and to the
dosing-interval mix, which is drawn uniformly over 6, 8 and 12 h here rather
than following the paper's dose-dependent interval rules. No gate is placed on
the trough band for that reason.

# Assumptions and deviations

- **Residual error.** Chen 2023 used a combined additive and *exponential*
  residual model, `Y = F * exp(eps1) + eps2`. rxode2 (tested on 5.1.7) cannot
  simulate a log-normal residual combined with an additive one -
  `Cc ~ lnorm(expSd) + add(addSd)` *parses and compiles*, then fails at
  `rxSolve()` with "cannot find additive standard deviation for 'Cc'; this
  parameter could be estimated by another endpoint, to fix move outside of
  error expression." Confirm it at the solve, not the build: a build-only
  check succeeds and makes this note look wrong. The exponential arm is
  therefore encoded as its
  first-order equivalent proportional arm, `Cc ~ add(addSd) + prop(propSd)`
  with `propSd = 0.177`, since `F * exp(eps1)` reduces to `F * (1 + eps1)` for
  small `eps1`. The cost is 0.8% on the residual magnitude: the paper's
  log-scale SD of 0.177 implies a linear-space CV of
  `sqrt(exp(0.177^2) - 1) = 0.1784` against the 0.177 encoded. The additive arm
  is carried unchanged at 0.154 mg/L.
- **Between-subject variance scale.** Table 3 reports the CL and V
  variabilities as 27.5% and 27.2% CV and, in the same block, their covariance
  as 0.0536 on the OMEGA scale. The paper does not state which %CV convention
  it used. Because the off-diagonal is given on the variance scale, the
  diagonals are read on the same scale here, `omega^2 = (%CV / 100)^2`, giving
  0.075625 and 0.073984 and an implied correlation of 0.717. Under the
  alternative exact log-normal reading, `omega^2 = log(1 + CV^2)`, the
  variances would be 0.072919 and 0.071392 and the correlation 0.743. The
  difference is under 4% in variance and changes nothing in any check above.
- **Albumin reference constant.** The Table 3 volume equation uses 36.2 in
  `(36.2 / ALB)^theta_ALB`, numerically identical to the TV(V) estimate of
  36.2 L that appears two lines above it. The Table 1 albumin median is
  36.3 g/L, so 36.2 g/L is a plausible rounded cohort median and the
  coincidence appears genuine. Were it instead a typesetting duplication and
  the true reference 36.3, the effect on V would be
  `(36.2/36.3)^0.279 = 0.9992`, i.e. under 0.1% - immaterial to every result
  here. The printed 36.2 is used.
- **eGFR is supplied uncapped.** The `CRCL` column takes the raw Equation 11
  value; the 120 mL/min/1.73 m^2 ceiling is applied inside `model()` because it
  is a fitted feature of the covariate model (Table 2 Model D), not a
  data-preparation step. Supplying an already-capped value is harmless; the
  `min()` is idempotent.
- **Unit conversions for eGFR.** Equation 11 takes height in metres, serum
  creatinine in mg/dL and BUN in mg/dL, whereas Chen 2023 Table 1 reports
  height in cm, creatinine in umol/L and BUN in mmol/L. A user computing
  `CRCL` from raw laboratory values must convert: cm / 100, umol/L / 88.4,
  mmol/L * 2.8.
- **Postmenstrual age is in weeks.** The canonical `PAGE` register default is
  months, but the register entry explicitly permits weeks for models whose
  source equations are written on that scale. TM50 = 37.0 weeks here, so this
  model declares and uses weeks throughout.
- **Representative covariates for Table 5.** The body weight and postmenstrual
  age used for each Chen 2023 age band are chosen in this vignette, not taken
  from the paper, which reports the recommendations by band rather than for a
  named reference patient. The eGFR band midpoints (45, 75, 105 and 140
  mL/min/1.73 m^2) are likewise chosen; 140 stands in for the open-ended
  ">120" band, where the model's cap makes any value above 120 equivalent.
  Where Table 5 gives a dose range, its midpoint is used.
- **Virtual-cohort covariate distributions.** Chen 2023 reports Table 1
  covariates marginally, not jointly. Weight and postmenstrual age are coupled
  here through a weight-for-age relationship so no 3 kg neonate is assigned an
  adolescent age, but eGFR, albumin, dosing interval and daily dose are drawn
  independently of age and of each other. The real cohort's correlations - a
  neonate is both smaller and more likely to be renally immature - are
  therefore not reproduced, which is one reason the simulated concentration
  distribution is wider than Table 1's.
- **The published concentrations come from clinical TDM sampling**, which is
  triggered by clinical concern rather than randomised, and their sampling
  windows are broad (footnote c). The comparison above is therefore between a
  randomly sampled virtual cohort and a clinically selected observed one, and
  is gated with headroom accordingly.
- **No external-validation cohort.** The 53-subject Shanghai Children's
  Medical Center cohort is described in the population metadata but not
  simulated; it contributed no parameter estimates.
- All parameter values come from the paper's own text and tables. No value was
  digitised from a figure, obtained by correspondence, or carried from an
  upstream model.
