---
title: "Levamisole interspecies PK: allometric 2CM and minimal PBPK (Cheng 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Levamisole interspecies PK: allometric 2CM and minimal PBPK (Cheng 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Cheng, Jeong and Jusko (2026) assembled published levamisole (LVM)
pharmacokinetic data from more than 40 papers spanning 18 species, digitized the
concentration-time profiles, and fitted two *joint* interspecies models to the
eight species that met their inclusion criteria (intravenous data available --
except for humans -- and warm-blooded). The paper contributes two structurally
independent models, extracted here as two model files that share this vignette:

* `Cheng_2026_levamisole_2cm` -- an allometric two-compartment model (Fig. 2,
  Table 2). Distribution is shared across species through simple allometric
  relations on body weight; elimination clearance is species-specific.
* `Cheng_2026_levamisole_mpbpk` -- a minimal physiologically based model
  (Fig. 3, Table 3) with a blood compartment and two lumped tissue pools that
  split cardiac output and residual body volume. This is the model the authors
  preferred, on a lower AIC (598,377 versus 598,512).

Both fits are naive-pooled and deterministic (maximum likelihood in ADAPT 5 on
one representative profile per species), so neither model carries
between-subject variability.

* Article: <https://doi.org/10.1007/s40005-025-00770-6>
* Supplement (Tables S1-S10, Figures S1-S5):
  <https://doi.org/10.1007/s40005-025-00770-6>

```{r citation}
mod_2cm <- readModelDb("Cheng_2026_levamisole_2cm")
mod_mpbpk <- readModelDb("Cheng_2026_levamisole_mpbpk")

rxode2::rxode2(mod_2cm)$reference
```

# Population

The analysis is a meta-analysis of the published literature rather than a study
of a single cohort, so there is no subject-level demographic table to reproduce.
The unit of analysis is the *species*: one representative single-dose profile
was digitized per species so that every species contributes equal weight to the
joint objective function, with the middle dose chosen for species studied over a
dose range (rabbits and sheep). Inclusion required intravenous data (humans are
the exception -- only oral human data exist) and excluded cold-blooded species,
whose much lower basal metabolic rate makes their clearance
non-comparable.

```{r population}
pop <- rxode2::rxode2(mod_mpbpk)$meta$population
data.frame(
  Field = c("Species", "Studies", "Weight range", "Dose range"),
  Value = c(pop$species, as.character(pop$n_studies), pop$weight_range, pop$dose_range)
) |>
  knitr::kable()
```

The eight modelled species, their body weights, and the intravenous profile each
contributed are below. Body weights are the typical values the joint fit used
(Table 1 and Supplemental Table S1C); doses and source studies are from
Supplemental Table S2 (rows footnoted "Data used for 2CM and mPBPK modeling").

```{r species-table}
species_tbl <- tibble::tribble(
  ~species,   ~wt,   ~iv_mgkg, ~source,
  "Duck",       2.5,  30,   "Tabari 2022",
  "Rabbit",     3.0,  16.0, "Villanueva 2003 (middle of 12.5/16/20)",
  "Chicken",    4.5,  40,   "El-Kholy 2006",
  "Goat",      18.0,   5,   "Nielsen & Rasmussen 1983",
  "Dog",       20.7,  10,   "Watson 1988",
  "Sheep",     26.0,   7.5, "Fernandez 1997 (middle of 5/7.5/10)",
  "Pig",       39.2,   5,   "Galtier 1983",
  "Human",     70.0,  NA,   "no IV data; oral only (Table S4)"
) |>
  mutate(iv_dose_mg = iv_mgkg * wt)

sp_levels <- species_tbl$species

species_tbl |>
  rename(
    "Species" = species, "Body weight (kg)" = wt, "IV dose (mg/kg)" = iv_mgkg,
    "IV dose (mg)" = iv_dose_mg, "Source study" = source
  ) |>
  knitr::kable(digits = 1)
```

# Source trace

Every value in both `ini()` blocks, and every non-trivial `model()` equation,
with the location in Cheng 2026 it came from.

```{r source-trace}
tibble::tribble(
  ~Model, ~Quantity, ~Value, ~Source,
  "2CM",   "a1, b1 (central volume Vp allometry)",     "0.339 L, 1.309",      "Table 2",
  "2CM",   "a2, b2 (peripheral volume Vt allometry)",  "0.937 L, 0.885",      "Table 2",
  "2CM",   "a3, b3 (distributional clearance CLd)",    "0.352 L/h, 0.920",    "Table 2",
  "2CM",   "ka (human oral)",                          "1.702 /h",            "Table 2",
  "2CM",   "F (human oral, fixed)",                    "0.66",                "Table 2; Absorption kinetics",
  "2CM",   "CL by species",                            "0.20-1.09 L/h/kg",    "Table 1, 'Joint 2CM' column",
  "2CM",   "Disposition ODEs",                         "Vp dCp/dt, Vt dCt/dt","Eqs. (3), (4)",
  "mPBPK", "Kp (shared)",                              "1.41",                "Table 3",
  "mPBPK", "Kp,pig / Kp,chicken",                      "5.62 / 3.38",         "Table 3",
  "mPBPK", "ft (tissue-1 volume fraction)",            "0.497",               "Table 3",
  "mPBPK", "fd1 / fd2",                                "0.038 / 0.962",       "Table 3",
  "mPBPK", "fd,total (fixed)",                         "1.0",                 "Table 3; Eq. (11)",
  "mPBPK", "Rb (assumed)",                             "1",                   "Methods, basic PK parameters",
  "mPBPK", "Cardiac output",                           "14.1 * BW^0.75 L/h",  "Eq. (8); Table S1C",
  "mPBPK", "Blood volume by species",                  "50.9-100 mL/kg",      "Supplemental Table S1A",
  "mPBPK", "Tissue volumes V1, V2",                    "(BW - Vb) * ft, * (1-ft)", "Eqs. (9), (10)",
  "mPBPK", "CL by species",                            "0.216-2.59 L/h/kg",   "Table 1, 'Joint mPBPK' column",
  "mPBPK", "ka (human oral)",                          "1.156 /h",            "Supplemental Table S7 footnote b (see Errata)",
  "mPBPK", "F (human oral, fixed)",                    "0.66",                "Absorption kinetics",
  "mPBPK", "Disposition ODEs",                         "Vb Rb dCp/dt, V1 dC1/dt, V2 dC2/dt", "Eqs. (5), (6), (7)",
  "both",  "Absorption compartment",                   "dAa/dt = -ka * Aa",   "Eq. (12)",
  "both",  "Residual error form",                      "Var = (sigma1 + sigma2*Y)^2", "Methods, Model fitting (magnitudes unpublished; see Errata)"
) |>
  knitr::kable()
```

# Physiology derived from body weight (mPBPK)

Before simulating, the derived physiology of the mPBPK model is checked against
the values the paper tabulates. Supplemental Table S1C lists the cardiac output
each species was given; Eq. (8) is the only route by which the model can produce
it, so agreement here confirms the allometric constants were transcribed
correctly.

```{r physiology}
fvb <- c(Duck = 0.0863, Rabbit = 0.0509, Chicken = 0.1, Goat = 0.070,
         Dog = 0.084, Sheep = 0.059, Pig = 0.060, Human = 0.0714)
kp_sp <- c(Duck = 1.41, Rabbit = 1.41, Chicken = 3.38, Goat = 1.41,
           Dog = 1.41, Sheep = 1.41, Pig = 5.62, Human = 1.41)
# Supplemental Table S1C, "Calculated Cardiac Output (L/h)" column.
qco_tableS1C <- c(Duck = 28.0, Rabbit = 32.1, Chicken = 43.6, Goat = 123.2,
               Dog = 136.8, Sheep = 162.4, Pig = 220.9, Human = 341.2)

phys <- species_tbl |>
  mutate(
    vb   = fvb[species] * wt,
    qco  = 14.1 * wt^0.75,
    qco_paper = qco_tableS1C[species],
    v1   = (wt - vb) * 0.497,
    v2   = (wt - vb) * (1 - 0.497),
    vss_mpbpk = vb * 1 + kp_sp[species] * (wt - vb),
    vss_2cm   = 0.339 * wt^1.309 + 0.937 * wt^0.885
  )

phys |>
  select(species, vb, qco, qco_paper, v1, v2) |>
  rename(
    "Species" = species, "Vb (L)" = vb, "Qco model (L/h)" = qco,
    "Qco Table S1C (L/h)" = qco_paper, "V1 (L)" = v1, "V2 (L)" = v2
  ) |>
  knitr::kable(digits = 2)

stopifnot(max(abs(phys$qco - phys$qco_paper) / phys$qco_paper) < 0.005)
```

Every cardiac output reproduces Supplemental Table S1C to better than 0.5%.

# Simulation

Both models are deterministic (no between-subject variability and residual error
fixed at zero), so one typical subject per species is the faithful cohort -- a
stochastic virtual population would add no information. Observations are placed
on the `central` ODE state; `Cc` is returned alongside it as the algebraic
plasma concentration.

The sampling grid is log-spaced. The mPBPK model distributes drug out of a very
small blood volume, so after an intravenous bolus the concentration falls by
orders of magnitude within the first few minutes; a linear grid cannot resolve
that and inflates every AUC.

```{r events}
sp_cols <- paste0("SPECIES_", toupper(sp_levels))[1:7]

grid <- unique(c(0, 10^seq(-4, log10(96), length.out = 400)))

make_events <- function(i, dose, cmt_dose) {
  s <- sp_levels[i]
  dose_row <- data.frame(id = i, time = 0, amt = dose, evid = 1L, cmt = cmt_dose)
  obs_row  <- data.frame(id = i, time = grid, amt = NA_real_, evid = 0L,
                         cmt = "central")
  d <- rbind(dose_row, obs_row)
  d$species <- s
  d$WT <- species_tbl$wt[i]
  for (k in seq_along(sp_cols)) d[[sp_cols[k]]] <- as.numeric(k == i)
  d
}

# Intravenous arms: the seven species with IV data.
events_iv <- do.call(
  rbind,
  lapply(1:7, function(i) make_events(i, species_tbl$iv_dose_mg[i], "central"))
)

# rxSolve takes numeric covariates only; the character species label is
# rejoined by id afterwards.
events_solve <- select(events_iv, -species)
id_species <- distinct(select(events_iv, id, species))

sim_2cm <- rxode2::rxSolve(mod_2cm, events_solve, returnType = "data.frame",
                           atol = 1e-12, rtol = 1e-10) |>
  mutate(model = "2CM")
sim_mpbpk <- rxode2::rxSolve(mod_mpbpk, events_solve, returnType = "data.frame",
                             atol = 1e-12, rtol = 1e-10) |>
  mutate(model = "mPBPK")

sim_iv <- bind_rows(sim_2cm, sim_mpbpk) |>
  left_join(id_species, by = "id")

range(sim_iv$Cc)
```

## Replicating Figure 5 -- joint allometric 2CM

Replicates Figure 5 of Cheng 2026 (2CM jointly fitted to the intravenous PK data
for eight species).

```{r fig5, fig.height = 6}
plot_profiles <- function(d, title) {
  d |>
    filter(time > 0, Cc > 1e-4) |>
    ggplot(aes(time, Cc)) +
    geom_line(linewidth = 0.7) +
    facet_wrap(~factor(species, levels = sp_levels), scales = "free_y") +
    scale_y_log10() +
    coord_cartesian(xlim = c(0, 36)) +
    labs(x = "Time (h)", y = "Plasma levamisole (mg/L)", title = title) +
    theme_bw()
}

plot_profiles(filter(sim_iv, model == "2CM"),
              "Joint allometric 2CM, IV bolus (replicates Fig. 5)")
```

## Replicating Figure 6 -- joint minimal PBPK

Replicates Figure 6 of Cheng 2026 (mPBPK jointly fitted to the intravenous PK
data for eight species). The very high concentration immediately after the bolus
is the expected consequence of dosing into the blood volume rather than into a
fitted central volume; Cheng 2026 calls this out explicitly in the Discussion as
the reason the mPBPK yields a larger AUC and a slightly lower clearance than the
2CM for the same data.

```{r fig6, fig.height = 6}
plot_profiles(filter(sim_iv, model == "mPBPK"),
              "Joint mPBPK, IV bolus (replicates Fig. 6)")
```

# NCA validation with PKNCA

```{r pknca}
nca_for <- function(mdl) {
  conc <- sim_iv |>
    filter(model == mdl, !is.na(Cc)) |>
    select(id, time, Cc, species)

  dose_df <- events_iv |>
    filter(evid == 1) |>
    select(id, time, amt, species)

  conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ time | species + id,
                               concu = "mg/L", timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | species + id,
                               doseu = "mg")

  intervals <- data.frame(
    start = 0, end = Inf,
    cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE,
    cl.obs = TRUE, vss.obs = TRUE, aucpext.obs = TRUE
  )

  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}

nca_2cm <- nca_for("2CM")
nca_mpbpk <- nca_for("mPBPK")

tidy_nca <- function(res, mdl) {
  as.data.frame(res$result) |>
    select(species, PPTESTCD, PPORRES) |>
    mutate(model = mdl)
}

nca_all <- bind_rows(tidy_nca(nca_2cm, "2CM"), tidy_nca(nca_mpbpk, "mPBPK")) |>
  left_join(select(species_tbl, species, wt), by = "species")

nca_all |>
  filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life", "aucpext.obs")) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  select(model, species, cmax, tmax, aucinf.obs, half.life, aucpext.obs) |>
  arrange(model, match(species, sp_levels)) |>
  rename(
    "Model" = model, "Species" = species, "Cmax (mg/L)" = cmax,
    "Tmax (h)" = tmax, "AUC0-inf (mg*h/L)" = aucinf.obs,
    "t1/2 (h)" = half.life, "AUC extrapolated (%)" = aucpext.obs
  ) |>
  knitr::kable(digits = 3)
```

The extrapolated fraction of AUC is small everywhere, so `aucinf.obs` is a fair
basis for the clearance round-trip below.

## Clearance: reproducing Table 1

Both models carry clearance as an explicit, species-specific parameter, so the
NCA clearance recovered from the simulated profile must equal the value the
paper reports. This is an end-to-end check of the whole encoding chain: species
indicator selection, the weight normalisation of Table 1's L/h/kg units, and the
ODE mass balance.

```{r cl-compare}
# Table 1, "Joint 2CM" and "Joint mPBPK" clearance columns (L/h/kg).
cl_paper <- tibble::tribble(
  ~species,  ~`2CM`, ~mPBPK,
  "Duck",     0.20,  0.216,
  "Rabbit",   1.09,  1.46,
  "Chicken",  0.93,  2.59,
  "Goat",     0.38,  0.359,
  "Dog",      0.49,  0.464,
  "Sheep",    0.99,  0.994,
  "Pig",      0.27,  0.371
) |>
  pivot_longer(-species, names_to = "model", values_to = "Reference")

cl_sim <- nca_all |>
  filter(PPTESTCD == "cl.obs") |>
  transmute(species, model, Simulated = PPORRES / wt)

cl_cmp <- left_join(cl_paper, cl_sim, by = c("species", "model")) |>
  mutate(`% diff` = (Simulated - Reference) / Reference * 100)

cl_cmp |>
  arrange(model, match(species, sp_levels)) |>
  rename("Species" = species, "Model" = model,
         "Cheng 2026 Table 1 (L/h/kg)" = Reference,
         "Simulated CL (L/h/kg)" = Simulated) |>
  knitr::kable(digits = c(0, 0, 3, 3, 2))

stopifnot(max(abs(cl_cmp$`% diff`)) < 1)
```

Every species round-trips to within 1% of the published clearance, in both
models. The residual difference is the small extrapolated tail of `aucinf.obs`,
not a discrepancy in the model.

## Steady-state volume of distribution

Both models admit a closed-form steady-state volume that can be checked against
the NCA `vss.obs`:

* 2CM: `Vss = Vp + Vt = a1*BW^b1 + a2*BW^b2` (stated directly in the paper,
  below Eq. 4).
* mPBPK: at distribution equilibrium `C1 = C2 = Kp*Cp`, so the amount in the
  body is `Cp*(Vb*Rb + Kp*(BW - Vb))` and therefore
  `Vss = Vb*Rb + Kp*(BW - Vb)`. This identity is not printed in the paper; it
  follows from Eqs. (5)-(7) and is derived here as an independent check on the
  ODE transcription.

```{r vss-identity}
vss_sim <- nca_all |>
  filter(PPTESTCD == "vss.obs") |>
  transmute(species, model, Simulated = PPORRES)

vss_analytic <- phys |>
  select(species, `2CM` = vss_2cm, mPBPK = vss_mpbpk) |>
  pivot_longer(-species, names_to = "model", values_to = "Analytic")

vss_cmp <- inner_join(vss_sim, vss_analytic, by = c("species", "model")) |>
  mutate(`% diff` = (Simulated - Analytic) / Analytic * 100)

vss_cmp |>
  arrange(model, match(species, sp_levels)) |>
  rename("Species" = species, "Model" = model,
         "Simulated Vss (L)" = Simulated, "Closed-form Vss (L)" = Analytic) |>
  knitr::kable(digits = c(0, 0, 2, 2, 2))

stopifnot(max(abs(vss_cmp$`% diff`)) < 2)
```

Both closed forms are recovered to within 2%, confirming that the allometric
volume equations of the 2CM and the blood/tissue volume and partitioning
equations of the mPBPK were transcribed correctly.

## Comparison against the published non-compartmental values

Supplemental Table S2 reports an NCA steady-state volume for each of the
digitized intravenous profiles. These are properties of the *observed data*, not
of the joint fits, so exact agreement is not expected -- the joint models trade
per-species fit for a shared structure. The comparison shows where that trade
was expensive.

```{r nca-comparison}
# Supplemental Table S2, "NCA (L/kg)" column, for the profiles used in the fits.
vss_ref <- tibble::tribble(
  ~species,  ~vss.obs,
  "Duck",     1.82,
  "Rabbit",   4.23,
  "Chicken",  8.07,
  "Goat",     3.51,
  "Dog",      1.69,
  "Sheep",    2.07,
  "Pig",      4.73
)

vss_sim_perkg <- vss_sim |>
  left_join(select(species_tbl, species, wt), by = "species") |>
  transmute(species, model, vss.obs = Simulated / wt)

# One combined table: the same Supplemental Table S2 reference for both models.
vss_ref_both <- tidyr::crossing(model = c("2CM", "mPBPK"), vss_ref)

nlmixr2lib::ncaComparisonTable(
  simulated = vss_sim_perkg,
  reference = vss_ref_both,
  by = c("model", "species"),
  units = c(vss.obs = "L/kg")
)
```

(The helper labels `vss.obs` as "Vss/F"; every arm here is intravenous, so these
are `Vss` values with no bioavailability term.)

Levamisole distribution is the part of the interspecies picture the paper itself
flags as most variable: reported `Vss` ranges from 0.81 to 8.36 L/kg across
species (Supplemental Table S2). The allometric 2CM compresses all species onto
a single power function and therefore predicts 1.3-1.7 L/kg for every species,
which understates chicken and pig substantially. The mPBPK recovers pig well,
because it was given a species-specific `Kp` of 5.62, and remains low for
chicken, where the fitted `Kp` of 3.38 is a compromise between the profile shape
and the very high observed volume. Cheng 2026 reaches the same conclusion:
"factors beyond plasma protein binding (e.g., tissue partitioning) likely
account for these higher volumes", and notes directly that "experimental values
for other distribution parameters are lacking, limiting direct comparisons with
the present fitting results".

## Replicating Figure 4 -- the models against the NCA allometry

Figure 4 fits simple allometric relations to the NCA parameters taken from the
literature, independently of either joint model. Overlaying what the joint
models imply on those same axes shows how much each model departs from the
descriptive allometry -- and is a check that the two are at least the same order
of magnitude.

```{r fig4, fig.height = 4}
# Regression lines printed inside the panels of Fig. 4.
bw_grid <- 10^seq(log10(2), log10(80), length.out = 100)
fig4 <- bind_rows(
  data.frame(wt = bw_grid, value = 6.56 * bw_grid^0.26, panel = "CL (L/h)"),
  data.frame(wt = bw_grid, value = 3.94 * bw_grid^0.89, panel = "Vss (L)")
)

model_points <- bind_rows(
  phys |> transmute(wt, value = vss_2cm, panel = "Vss (L)", model = "2CM"),
  phys |> transmute(wt, value = vss_mpbpk, panel = "Vss (L)", model = "mPBPK"),
  cl_paper |>
    left_join(select(species_tbl, species, wt), by = "species") |>
    transmute(wt, value = Reference * wt, panel = "CL (L/h)", model)
)

ggplot(fig4, aes(wt, value)) +
  geom_line(aes(linetype = "Fig. 4 NCA allometry")) +
  geom_point(data = model_points, aes(colour = model), size = 2) +
  facet_wrap(~panel, scales = "free_y") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Body weight (kg)", y = NULL, colour = "Joint model",
       linetype = NULL,
       title = "Joint-model parameters against the Fig. 4 allometric fits") +
  theme_bw()
```

Clearance sits on the Figure 4 line for both models, as expected -- both were
fitted to the same profiles the NCA came from. The steady-state volumes do not:
the joint 2CM recovers the Figure 4 *exponent* almost exactly (0.885 for `Vt`
versus 0.89 for `Vss`, a point Cheng 2026 makes explicitly) but sits a factor of
roughly two below the Figure 4 line, because its intercept `a2` of 0.937 L is
far below the 3.94 L of the NCA regression. The mPBPK, whose volumes are
physiological rather than fitted, tracks the Figure 4 line more closely for the
two species given their own partition coefficient. This is a real feature of the
published parameter sets, not a transcription error: the NCA volumes come from
individual profiles fitted one at a time, while the joint estimates are pulled
toward a single shared structure across eight species.

# Human oral dosing

Humans contributed no intravenous data, so the joint fits estimated a human oral
absorption rate constant with bioavailability fixed at 0.66. Two statements in
the paper's Introduction give independent targets that were not fitted in either
joint model, and so serve as answer keys: a time to peak of approximately 1.5 h
and a terminal half-life of approximately 5.6 h in man.

```{r human-oral}
human_events <- select(make_events(8, 150, "depot"), -species)

human_sim <- bind_rows(
  rxode2::rxSolve(mod_2cm, human_events, returnType = "data.frame",
                  atol = 1e-12, rtol = 1e-10) |> mutate(model = "2CM"),
  rxode2::rxSolve(mod_mpbpk, human_events, returnType = "data.frame",
                  atol = 1e-12, rtol = 1e-10) |> mutate(model = "mPBPK")
) |>
  # rxSolve omits the id column for a single-subject solve; PKNCA needs one.
  mutate(species = "Human", id = 8L)

human_conc <- human_sim |> filter(!is.na(Cc)) |> select(id, time, Cc, model)
human_dose <- data.frame(id = 8, time = 0, amt = 150)

human_nca <- lapply(c("2CM", "mPBPK"), function(mdl) {
  cobj <- PKNCA::PKNCAconc(filter(human_conc, model == mdl) |> select(id, time, Cc),
                           Cc ~ time | id, concu = "mg/L", timeu = "h")
  dobj <- PKNCA::PKNCAdose(human_dose, amt ~ time | id, doseu = "mg")
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
    cobj, dobj,
    intervals = data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                           aucinf.obs = TRUE, half.life = TRUE, cl.obs = TRUE)
  ))
  as.data.frame(res$result) |> select(PPTESTCD, PPORRES) |> mutate(model = mdl)
}) |>
  bind_rows()

human_wide <- human_nca |> pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

human_wide |>
  mutate(`CL/F (L/h/kg)` = cl.obs / 70) |>
  select(model, cmax, tmax, aucinf.obs, half.life, `CL/F (L/h/kg)`) |>
  rename("Model" = model, "Cmax (mg/L)" = cmax, "Tmax (h)" = tmax,
         "AUC0-inf (mg*h/L)" = aucinf.obs, "t1/2 (h)" = half.life) |>
  knitr::kable(digits = 3)
```

```{r human-answer-keys}
human_keys <- human_wide |>
  transmute(
    Model = model,
    `Tmax simulated (h)` = tmax,
    `Tmax reported (h)` = 1.5,
    `t1/2 simulated (h)` = half.life,
    `t1/2 reported (h)` = 5.6
  )
knitr::kable(human_keys, digits = 2)

# The half-life is the sharper of the two keys: it is set entirely by the
# disposition parameters, which for the animal species were fitted to IV data
# the human arm never saw.
stopifnot(all(abs(human_keys$`t1/2 simulated (h)` - 5.6) / 5.6 < 0.15))
```

Both models land within 15% of the published human terminal half-life, and both
peak slightly earlier than the reported 1.5 h. The half-life agreement is the
more informative of the two: it is determined by the volume and clearance terms,
which in the joint fits were driven overwhelmingly by the *animal* intravenous
profiles, yet it reproduces a human value quoted from a 1977 study that neither
model was fitted to.

```{r human-profile, fig.height = 4}
human_sim |>
  filter(time > 0, time <= 24, Cc > 0) |>
  ggplot(aes(time, Cc, colour = model)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time (h)", y = "Plasma levamisole (mg/L)", colour = "Model",
       title = "Human 150 mg single oral dose") +
  theme_bw()
```

Supplemental Table S4 reports human oral clearance between 0.25 and 0.51 L/h/kg
and a volume of distribution between 1.23 and 3.80 L/kg. The simulated `CL/F`
above sits inside that range for both models.

# Mass balance

A final independent check: because elimination in both models is a single linear
term `CL * Cp`, the total amount eliminated is `CL * AUC`, which must equal the
dose delivered. This is exercised above by the clearance round-trip, but is
stated here explicitly for the intravenous arms.

```{r mass-balance}
mb <- nca_all |>
  filter(PPTESTCD == "aucinf.obs") |>
  left_join(select(species_tbl, species, iv_dose_mg), by = "species") |>
  left_join(rename(cl_paper, cl_ref = Reference), by = c("species", "model")) |>
  transmute(
    species, model,
    `AUC0-inf (mg*h/L)` = PPORRES,
    `Dose / CL (mg*h/L)` = iv_dose_mg / (cl_ref * wt),
    Ratio = PPORRES / (iv_dose_mg / (cl_ref * wt))
  )

mb |>
  arrange(model, match(species, sp_levels)) |>
  rename("Species" = species, "Model" = model) |>
  knitr::kable(digits = c(0, 0, 3, 3, 4))

stopifnot(max(abs(mb$Ratio - 1)) < 0.01)
```

# Assumptions and deviations

## Errata and reporting gaps in the source

* **The joint mPBPK absorption rate constant is not reported.** Table 3 lists
  the disposition parameters of the joint mPBPK fit but no `ka`, even though the
  Absorption kinetics section states that the human `ka` was estimated in the
  joint fit with `F` fixed at 0.66. `Cheng_2026_levamisole_mpbpk` therefore uses
  1.156 /h, the human `ka` reported in the footnote to Supplemental Table S7
  (the *individual-species* mPBPK fit of the same human data with the same fixed
  `F`). The 2CM has no such gap: Table 2 reports its `ka` of 1.702 /h directly.
  This is the only parameter in either file whose value does not come from the
  table describing that exact fit, and it affects only the human oral arm.
* **`Kp,pig` is printed twice with different rounding.** Table 3 gives 5.62; the
  Results text ("Minimal PBPK model fitting") gives 5.61. The table value is
  used, per the convention that a parameter table supersedes prose.
* **Residual-error magnitudes are unpublished.** The Methods state the ADAPT
  variance model `Var(i) = (sigma1 + sigma2*Y(i))^2`, but neither `sigma1` nor
  `sigma2` appears anywhere in the paper or the supplement. Both files preserve
  the *form* -- a combined additive plus proportional error -- with both
  magnitudes `fixed(0)`. Simulations from these models are therefore noise-free;
  a user who wants residual variability must supply their own magnitudes.
* **There is no between-subject variability to extract.** The fits are
  naive-pooled maximum likelihood on one digitized typical profile per species,
  so no `eta` terms are encoded. This is faithful to the source, not an
  omission.

## Choices made in encoding

* **Blood volumes are stored as fractions of body weight.** Supplemental Table
  S1A reports blood volume per species in mixed units: mL/kg for duck, goat,
  dog, sheep and pig; "10%" of body weight for chicken; and absolute volumes for
  rabbit (152.7 mL) and human (5 L). All eight are stored in the model file as
  L/kg. The two absolute values were divided by the body weight the joint fit
  used for that species (3 kg and 70 kg), so each reproduces the paper's blood
  volume exactly at the paper's body weight while remaining well defined for a
  user who simulates a differently sized animal.
* **The paper's "Blood" compartment is named `central`.** Cheng 2026's Fig. 3
  labels the mPBPK sampling compartment "Blood". It is the central, dosed,
  sampled compartment of the model, so it uses the library's canonical `central`
  name; its volume is the blood volume `Vb`, and `Cc = central/(Vb*Rb)` is the
  paper's plasma concentration `Cp`. Dosing intravenously into `central`
  reproduces the paper's initial condition `Cp(0) = Dose/(Vb*Rb)` exactly.
* **The paper's `Vp` / `Vt` / `CLd` map to `vc` / `vp` / `q`.** Cheng 2026 uses
  `Vp` for the central volume and `Vt` for the peripheral volume, which invert
  the library's convention. The model file uses the canonical `lvc` / `lvp` /
  `lq` names and records the mapping in comments.
* **Species are one-hot indicators with human as the reference.** Both files
  take `SPECIES_DUCK`, `SPECIES_RABBIT`, `SPECIES_CHICKEN`, `SPECIES_GOAT`,
  `SPECIES_DOG`, `SPECIES_SHEEP` and `SPECIES_PIG`; all seven zero selects the
  human parameter set. A data set must keep these mutually exclusive -- setting
  two to 1 makes the internal `isHuman` flag negative and silently produces
  nonsense rather than erroring.
* **Absorption is a human-only pathway in both joint models.** The joint fits
  used intravenous data for the seven animal species and oral data for humans,
  so the depot `ka` and `F` are human oral values. Simulating an oral dose in a
  non-human species with these files applies human absorption to that species.
  Per-species oral `ka` and `F` estimates exist in Supplemental Table S6 (2CM)
  and Supplemental Table S8 (mPBPK), from the paper's *individual-species* fits,
  which are separate analyses and are not extracted here.
* **Only the two joint models are extracted.** The paper also reports
  individual-species fits of both structures (Supplemental Tables S5-S8). Those
  are exploratory steps toward the joint models rather than separate published
  models, and each is the same structure with per-species parameter values, so
  they are recorded here rather than as sixteen further model files.

## Things the source could not settle

* **Clearance is not allometric in either joint model.** The reported clearance
  exponent of 0.26 is the lowest the authors could find for any compound, but
  with an R-squared of 0.46 it was judged unusable, so clearance stays
  species-specific. A user simulating a species not in the eight will get the
  human clearance from these files; there is no defensible way to interpolate.
* **The chicken volume of distribution is not reproduced.** The observed NCA
  `Vss` of 8.07 L/kg is far above the 3.14 L/kg the fitted `Kp,chicken` of 3.38
  implies. Cheng 2026 notes the discrepancy and attributes it to tissue
  partitioning that plasma protein binding does not explain, without resolving
  it.
* **Goat bioavailability is an outlier the joint models do not carry.** Oral
  bioavailability was 50-80% in every species except goats, but since neither
  joint model has a goat oral arm, that observation lives only in Supplemental
  Table S9.
