---
title: "Deoxypodophyllotoxin interspecies PBPK (Chen 2016)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deoxypodophyllotoxin interspecies PBPK (Chen 2016)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r citation, echo = FALSE, results = "asis"}
cat(
  "* Citation: ",
  rxode2::rxode(readModelDb("Chen_2016_deoxypodophyllotoxin_rat_pbpk"))$reference,
  "\n"
)
```

* Article: [Front Pharmacol. 2016;7:488](https://doi.org/10.3389/fphar.2016.00488)
  (open access)

Deoxypodophyllotoxin (DPT) is a pre-clinical anti-tumour candidate. Chen and
colleagues built a whole-body physiologically based pharmacokinetic (PBPK)
model of DPT from *in vitro* and *in silico* inputs, developed it in the rat,
extrapolated it to mouse, monkey and dog, validated each against in-vivo
plasma (and, in the mouse, tissue) data, and then used it to make a
first-in-human prediction. They made a second, independent human prediction by
interspecies allometric scaling.

This paper therefore contributes **six** models to `nlmixr2lib`.

```{r model-list, echo = FALSE}
dpt_models <- c(
  "Chen_2016_deoxypodophyllotoxin_mouse_pbpk",
  "Chen_2016_deoxypodophyllotoxin_rat_pbpk",
  "Chen_2016_deoxypodophyllotoxin_monkey_pbpk",
  "Chen_2016_deoxypodophyllotoxin_dog_pbpk",
  "Chen_2016_deoxypodophyllotoxin_human_pbpk",
  "Chen_2016_deoxypodophyllotoxin_human_allometric"
)
data.frame(
  Model = dpt_models,
  Role = c(
    "PBPK, mouse (also the only tissue-level validation)",
    "PBPK, rat -- the reference species the others are derived from",
    "PBPK, cynomolgus monkey",
    "PBPK, beagle dog",
    "PBPK, human -- forward prediction, 16 mg IV",
    "Two-compartment human model from Dedrick-plot allometric scaling"
  )
) |>
  knitr::kable(caption = "The six models contributed by Chen 2016.")
```

The five PBPK files share one structure and differ only in the physiological
and metabolic constants of their species. They are kept as separate files
rather than one covariate-switched file because the authors ran them as five
separate parameterisations, each drawing its physiology from a different
literature source (Table 1 footnotes a-g).

# Population

The in-vivo data are four single-dose intravenous animal studies.

```{r population, echo = FALSE}
pop_tbl <- do.call(rbind, lapply(dpt_models[1:5], function(m) {
  p <- readModelDb(m)()$population
  data.frame(
    Model = sub("^Chen_2016_deoxypodophyllotoxin_", "", m),
    Species = p$species,
    N = p$n_subjects,
    Weight = p$weight_range %||% NA_character_,
    Doses = p$dose_range
  )
}))
knitr::kable(
  pop_tbl,
  row.names = FALSE,
  caption = "Study populations (Methods, 'In vivo Pharmacokinetic Study')."
)
```

The mouse, monkey and dog studies were run for this paper. The rat plasma
profiles were taken from the authors' earlier publication (Liu et al. 2016),
so Chen 2016 does not report the rat cohort size, age or sampling schedule.
There are **no human data at all** -- the human dose of 16 mg was chosen from
the monkey maximum tolerated dose of 4 mg/kg, and both human models are
forward predictions.

The same information is available programmatically via
`readModelDb("Chen_2016_deoxypodophyllotoxin_dog_pbpk")()$population`.

# Source trace

Every value in the six model files comes from one of the locations below.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Quantity, ~`Source location`,
  "Organ volumes, blood flows, Kt:pl for all 5 species", "Table 1 (per-species columns)",
  "Brain permeability-surface product PS", "Table 1, 'PS (mL/min)' row",
  "Brain vascular / extravascular volume split (3% / 97%)", "Table 1 footnote i",
  "Liver blood flow = hepatic artery + GI tract + spleen", "Table 1 footnote h",
  "Rest-of-body volume and flow by subtraction; Kt:pl,rest = 0.01", "Table 1 footnote j",
  "Vein / artery = 2/3 and 1/3 of blood volume", "Table 1 footnote k",
  "PBSF, Km and Vmax for M2 and M7, Hill coefficient gamma", "Table 2 (per-species columns)",
  "Microsomal protein yield (44.8 / 48.8 / 77.9 mg/g liver)", "Table 2 footnote",
  "Fraction unbound in plasma, all 5 species", "Results, 'Plasma Protein Binding of DPT in Five Species'",
  "Perfusion-limited tissue ODE", "Methods, 'PBPK Model Development', eq. for non-elimination tissues",
  "Arterial, venous and lung blood ODEs", "Methods, 'PBPK Model Development'",
  "Liver ODE including the PBSF x (CL_M2 + CL_M7) sink", "Methods, 'PBPK Model Development'",
  "Permeability-limited brain ODEs and the PS allometric scaling", "Results, 'PBPK Model Development and Validation'",
  "CL_M2 Michaelis-Menten sum and CL_M7 Hill equation", "Results, equations following the brain model",
  "Rbp = 1", "Methods, 'PBPK Model Development' (citing Poulin and Theil 2002)",
  "Kt:pl = Kt:pl,rat x fu / fu,rat", "Results, 'PBPK Model Development and Validation'",
  "Human biexponential C = 0.628 e^-0.147t + 0.066 e^-0.011t", "Results, 'Interspecies Allometric Scaling'",
  "Allometric regressions CL = 44.28 W^0.832, Vss = 1.99 W^0.857", "Results, 'Interspecies Allometric Scaling'",
  "Predicted and observed NCA parameters used as targets below", "Table 3"
) |>
  knitr::kable(caption = "Source location for every model equation and parameter.")
```

Two quantities are **not** printed in the paper and are flagged here and in the
model files:

* **Molecular weight, 398.4 g/mol.** A chemical constant for the DPT formula
  C22H22O7 (PubChem CID 73435), not a fitted parameter. It is needed purely as
  a unit bridge: the ODEs run on a mass concentration scale while the Table 2
  Michaelis constants are in uM.
* **Residual error and inter-individual variances.** The paper's visual
  predictive checks fitted a proportional residual error and exponential
  inter-individual variability on hepatic blood flow and metabolic velocity in
  Phoenix NLME, but reports none of the estimates. They are encoded as
  `fixed(0)` rather than invented.

# Model structure

```{r structure, echo = FALSE}
cat(readModelDb("Chen_2016_deoxypodophyllotoxin_mouse_pbpk")()$description)
```

Eleven perfusion-rate-limited organs plus arterial and venous blood. The
gastrointestinal tract and spleen drain into the liver, not into venous blood,
so the liver receives hepatic-artery, portal and splenic inflow. The brain is
the one exception to perfusion-rate limitation: a preliminary perfusion-limited
brain over-predicted mouse brain concentrations, so the authors split it into a
vascular and an extravascular sub-compartment coupled by a
permeability-surface product.

```{r show-model}
mod_mouse <- rxode2::rxode(readModelDb("Chen_2016_deoxypodophyllotoxin_mouse_pbpk"))
mod_rat <- rxode2::rxode(readModelDb("Chen_2016_deoxypodophyllotoxin_rat_pbpk"))
mod_monkey <- rxode2::rxode(readModelDb("Chen_2016_deoxypodophyllotoxin_monkey_pbpk"))
mod_dog <- rxode2::rxode(readModelDb("Chen_2016_deoxypodophyllotoxin_dog_pbpk"))
mod_human <- rxode2::rxode(readModelDb("Chen_2016_deoxypodophyllotoxin_human_pbpk"))
mod_allom <- rxode2::rxode(readModelDb("Chen_2016_deoxypodophyllotoxin_human_allometric"))

mod_mouse$state
```

# Structural checks on the transcribed constants

Before simulating anything, four internal consistency relations that the paper
states in prose or footnotes are checked against the transcribed `ini()`
values. Each would fail loudly if a Table 1 or Table 2 column had been read off
by one.

```{r theta-helper}
# Named vector of the fixed-effect values of a packaged model.
dpt_theta <- function(model_name) {
  rxode2::rxode(readModelDb(model_name))$theta
}

dpt_kp <- function(model_name) {
  th <- dpt_theta(model_name)
  lkp <- th[grepl("^lkp_", names(th))]
  stats::setNames(exp(lkp), sub("^lkp_", "", names(lkp)))
}

th <- lapply(
  stats::setNames(dpt_models[1:5], c("mouse", "rat", "monkey", "dog", "human")),
  dpt_theta
)
kp <- lapply(
  stats::setNames(dpt_models[1:5], c("mouse", "rat", "monkey", "dog", "human")),
  dpt_kp
)
```

## 1. Kt:pl is the rat value rescaled by the unbound fraction

The paper's central assumption is that the *unbound* tissue-to-plasma ratio is
identical across species, so `Kt:pl = Kt:pl,rat * fu / fu,rat` for every tissue
of every species (Results, "PBPK Model Development and Validation").

Table 1 prints Kt:pl to two decimal places (four for the rest-of-body column),
and the rat column it is all derived from carries that same rounding, so the
correct criterion is agreement to within **one unit in the last printed
place** -- not a percentage. A transposed or mis-read column would be out by
far more than one digit.

```{r check-kp}
fu_all <- vapply(th, function(x) unname(x[["fu"]]), numeric(1))
kp_rat <- kp$rat

kp_check <- do.call(rbind, lapply(names(kp), function(sp) {
  expected <- kp_rat * fu_all[[sp]] / fu_all[["rat"]]
  printed <- kp[[sp]][names(kp_rat)]
  # The last printed place in Table 1: 1e-4 for the rest-of-body column,
  # 1e-2 everywhere else.
  ulp <- ifelse(printed < 0.01, 1e-4, 1e-2)
  data.frame(
    species = sp,
    tissue = names(kp_rat),
    printed = unname(printed),
    expected = unname(expected),
    dev_ulp = unname(abs(printed - expected) / ulp)
  )
}))

stopifnot(all(kp_check$dev_ulp < 1))
stopifnot(stats::median(kp_check$dev_ulp) < 0.2)

cat(sprintf(
  paste0(
    "All %d Kt:pl values across 5 species reproduce Kt:pl,rat * fu / fu_rat\n",
    "to within one unit in the last printed place (median %.2f, max %.2f).\n"
  ),
  nrow(kp_check), stats::median(kp_check$dev_ulp), max(kp_check$dev_ulp)
))
```

The handful of cells above half a unit are listed below. Most are the muscle
row: reproducing the mouse, monkey, dog and human muscle Kt:pl exactly requires
a rat muscle Kt:pl near 0.76 rather than the 0.75 Table 1 prints, which is
simply the rounding of the rat reference column propagating outward.

```{r check-kp-detail}
kp_check[kp_check$dev_ulp > 0.5, ] |>
  dplyr::mutate(
    printed = round(printed, 5),
    expected = round(expected, 5),
    dev_ulp = round(dev_ulp, 2)
  ) |>
  knitr::kable(
    row.names = FALSE,
    caption = "Kt:pl cells more than half a printed digit from the fu-scaling rule."
  )
```

## 2. PBSF is the microsomal protein yield times the liver weight

Table 2's footnote defines PBSF as the product of the microsomal protein yield
and the liver weight from Table 1. Yields are 44.8 mg/g liver for mouse and
rat, 48.8 for monkey and human, and 77.9 for dog.

```{r check-pbsf}
mppgl <- c(mouse = 44.8, rat = 44.8, monkey = 48.8, dog = 77.9, human = 48.8)
# Liver volume is in mL for the animals and L for human; assume density 1 g/mL.
liver_g <- vapply(
  names(th),
  function(sp) unname(th[[sp]][["v_liver"]]) * if (sp == "human") 1000 else 1,
  numeric(1)
)
pbsf_printed <- vapply(th, function(x) unname(x[["pbsf"]]), numeric(1))
pbsf_expected <- mppgl[names(pbsf_printed)] * liver_g[names(pbsf_printed)]

stopifnot(all(abs(pbsf_printed - pbsf_expected) < 0.01))
knitr::kable(
  data.frame(
    Species = names(pbsf_printed),
    `Yield (mg/g)` = unname(mppgl[names(pbsf_printed)]),
    `Liver (g)` = unname(liver_g[names(pbsf_printed)]),
    `PBSF printed` = unname(pbsf_printed),
    `PBSF = yield x liver` = unname(pbsf_expected),
    check.names = FALSE
  ),
  caption = "Table 2 PBSF reproduced exactly from its own footnote."
)
```

## 3. PS scales allometrically from the fitted mouse value

`PS_i = PS_mouse * (W_i / W_mouse)^0.67`, with PS estimated only in the mouse.

```{r check-ps}
bw_all <- c(mouse = 0.02, rat = 0.25, monkey = 4, dog = 8.5, human = 70)
ps_printed <- vapply(th, function(x) unname(x[["ps_brain"]]), numeric(1))
ps_expected <- unname(ps_printed[["mouse"]]) *
  (bw_all[names(ps_printed)] / 0.02)^0.67
ps_ratio <- ps_printed / ps_expected

# PS_mouse is printed to only two significant figures (0.0024), and that
# rounding propagates to every scaled species -- hence a relative rather than
# absolute tolerance. The ratios are all ~1.02, consistent with an unrounded
# PS_mouse of about 0.00244.
stopifnot(all(abs(ps_ratio - 1) < 0.03))
knitr::kable(
  data.frame(
    Species = names(ps_printed),
    `Body weight (kg)` = unname(bw_all[names(ps_printed)]),
    `PS printed (mL/min)` = unname(ps_printed),
    `PS scaled from mouse` = round(unname(ps_expected), 5),
    Ratio = round(unname(ps_ratio), 3),
    check.names = FALSE
  ),
  caption = "Table 1 PS row reproduced by the 0.67-power allometric rule."
)
```

## 4. Venous return balances cardiac output

Every organ that drains to venous blood, plus the liver (which collects the
gastrointestinal and splenic flows), must sum to the cardiac output.

```{r check-flow}
venous_organs <- c(
  "adipose", "muscle", "kidney", "brain", "heart", "skin", "other", "liver"
)
flow_check <- do.call(rbind, lapply(names(th), function(sp) {
  x <- th[[sp]]
  inflow <- sum(vapply(venous_organs, function(o) unname(x[[paste0("q_", o)]]), numeric(1)))
  data.frame(
    species = sp,
    cardiac_output = unname(x[["q_lung"]]),
    venous_inflow = inflow,
    pct_diff = 100 * (inflow / unname(x[["q_lung"]]) - 1)
  )
}))

# Mouse, monkey, dog and human balance to the printed precision. The rat
# column of Table 1 is internally inconsistent by 0.83 mL/min (1.0%) -- see
# the Errata; it is transcribed as printed, not adjusted.
stopifnot(all(abs(flow_check$pct_diff) < 1.1))
knitr::kable(
  flow_check,
  row.names = FALSE,
  digits = c(0, 2, 2, 2),
  caption = "Venous inflow versus cardiac output, per species."
)
```

# Simulation

All simulations are deterministic typical-value solves: the models carry no
random effects, so a single subject per dose group is the whole cohort. The
observation grid is dense in the first two minutes because an intravenous bolus
into a small venous blood pool produces a short, sharp mixing spike whose area
is a real part of the AUC.

```{r sim-helper}
# Observation grid: dense over the bolus mixing spike, coarser later.
dpt_grid <- function(tmax) {
  sort(unique(c(
    seq(0, 2, by = 0.005),
    seq(2, 30, by = 0.1),
    seq(30, tmax, by = 1)
  )))
}

# Solve one species at one dose. `cmt = "venous"` names the ODE state directly
# on both the dose and the observation rows.
dpt_solve <- function(model, dose_amount, tmax, treatment, id) {
  ev <- rxode2::et(amt = dose_amount, cmt = "venous") |>
    rxode2::et(dpt_grid(tmax), cmt = "venous")
  out <- as.data.frame(rxode2::rxSolve(model, ev, returnType = "data.frame"))
  out$treatment <- treatment
  out$id <- id
  out$amt_dosed <- dose_amount
  out
}
```

```{r sim-arms}
# Body weights are the Table 1 reference weights, since the Table 1 physiology
# is tabulated for those weights.
arms <- tibble::tribble(
  ~species, ~dose_mgkg, ~bw, ~tmax,
  "mouse", 12.5, 0.02, 240,
  "mouse", 25.0, 0.02, 240,
  "rat", 1.0, 0.25, 240,
  "rat", 2.0, 0.25, 240,
  "rat", 4.0, 0.25, 240,
  "monkey", 0.5, 4.00, 240,
  "monkey", 1.0, 4.00, 240,
  "monkey", 2.0, 4.00, 240
) |>
  dplyr::mutate(
    # Animal models use ug and mL, so a mg/kg dose becomes ug.
    dose_amount = dose_mgkg * bw * 1000,
    treatment = paste0(species, " ", dose_mgkg, " mg/kg"),
    id = dplyr::row_number()
  )

models <- list(
  mouse = mod_mouse, rat = mod_rat, monkey = mod_monkey,
  dog = mod_dog, human = mod_human
)

sim_animals <- dplyr::bind_rows(lapply(seq_len(nrow(arms)), function(i) {
  a <- arms[i, ]
  dpt_solve(models[[a$species]], a$dose_amount, a$tmax, a$treatment, a$id)
}))

# The dog is dosed once and sampled to 840 min, the paper's sampling window.
sim_dog <- dpt_solve(mod_dog, 0.3 * 8.5 * 1000, 840, "dog 0.3 mg/kg", 9L)

# The human PBPK model works in mg and L, so 16 mg is the dose as given.
sim_human <- dpt_solve(mod_human, 16, 240, "human 16 mg (PBPK)", 10L)

nrow(sim_animals) + nrow(sim_dog) + nrow(sim_human)
```

## Replicating Figure 4

```{r fig4-animals, fig.cap = "Replicates Figures 4A-D of Chen 2016: predicted DPT plasma concentration-time profiles in rat, mouse, monkey and dog after intravenous dosing."}
plot_dat <- dplyr::bind_rows(sim_animals, sim_dog) |>
  dplyr::mutate(
    species = factor(
      sub(" .*$", "", treatment),
      levels = c("rat", "mouse", "monkey", "dog")
    )
  ) |>
  dplyr::filter(time > 0, Cc > 0)

ggplot2::ggplot(plot_dat, ggplot2::aes(time, Cc, colour = treatment)) +
  ggplot2::geom_line() +
  ggplot2::facet_wrap(~species, scales = "free") +
  ggplot2::scale_y_log10() +
  ggplot2::labs(
    x = "Time (min)",
    y = "Plasma DPT (ug/mL)",
    colour = NULL
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

```{r fig4e, fig.cap = "Replicates Figure 4E of Chen 2016: the two independent human predictions, whole-body PBPK and Dedrick-plot allometric scaling, after a 16 mg intravenous bolus."}
sim_allom <- {
  ev <- rxode2::et(amt = 16, cmt = "central") |>
    rxode2::et(dpt_grid(240), cmt = "central")
  out <- as.data.frame(rxode2::rxSolve(mod_allom, ev, returnType = "data.frame"))
  out$treatment <- "human 16 mg (allometric)"
  out$id <- 11L
  out$amt_dosed <- 16
  out
}

dplyr::bind_rows(sim_human, sim_allom) |>
  dplyr::filter(time > 0, Cc > 0) |>
  ggplot2::ggplot(ggplot2::aes(time, Cc, colour = treatment)) +
  ggplot2::geom_line() +
  ggplot2::scale_y_log10() +
  ggplot2::labs(x = "Time (min)", y = "Plasma DPT (ug/mL)", colour = NULL) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

The allometric model is an exact algebraic re-parameterisation of the
biexponential the paper prints, so it can be checked against that equation to
machine precision rather than merely by eye.

```{r check-allometric}
published_biexp <- 0.628 * exp(-0.147 * sim_allom$time) +
  0.066 * exp(-0.011 * sim_allom$time)
max_rel_diff <- max(abs(sim_allom$Cc / published_biexp - 1))

stopifnot(max_rel_diff < 1e-3)
cat(sprintf(
  "Allometric model versus the printed C = 0.628 e^-0.147t + 0.066 e^-0.011t:\n  maximum relative difference %.2e over 0-240 min.\n",
  max_rel_diff
))
```

## Mass balance

The intended sink in the PBPK models is hepatic metabolism, which the model
exposes as `rate_metabolism`. Total body amount at any time should therefore
equal the dose minus the integral of that rate. This exercises the dose
placement, every inter-organ flow term and the unit bridge in the metabolic arm
at once.

There is a second sink, and it is not intentional. Because Table 1's rat column
does not balance (check 4 above), the rat venous pool receives 83.07 mL/min of
inflow while discharging 83.90 mL/min to the lung, so the transcribed rat
physiology destroys drug at a rate of `(qc - venous inflow) * C_venous`. That
is a property of the published table, not of the transcription, and the values
are encoded as printed rather than adjusted to close the balance. Accounting
for it explicitly turns the defect into a quantitative check.

```{r mass-balance}
dpt_states <- mod_mouse$state

dpt_cumtrap <- function(tt, y) {
  c(0, cumsum(diff(tt) * (utils::head(y, -1) + utils::tail(y, -1)) / 2))
}

dpt_mass_balance <- function(sim_one, leak_flow) {
  total <- rowSums(sim_one[, dpt_states, drop = FALSE])
  tt <- sim_one$time
  dose <- sim_one$amt_dosed[1]
  cum_met <- dpt_cumtrap(tt, sim_one$rate_metabolism)
  cum_leak <- dpt_cumtrap(tt, leak_flow * sim_one$Cc)
  c(
    metabolism_only = max(abs((total + cum_met) / dose - 1)),
    with_flow_imbalance = max(abs((total + cum_met + cum_leak) / dose - 1))
  )
}

# Positive = cardiac output exceeds the summed venous inflow, i.e. a leak.
leak <- stats::setNames(
  flow_check$cardiac_output - flow_check$venous_inflow,
  flow_check$species
)

mb <- rbind(
  mouse = dpt_mass_balance(dplyr::filter(sim_animals, id == 1), leak[["mouse"]]),
  rat = dpt_mass_balance(dplyr::filter(sim_animals, id == 3), leak[["rat"]]),
  monkey = dpt_mass_balance(dplyr::filter(sim_animals, id == 6), leak[["monkey"]]),
  dog = dpt_mass_balance(sim_dog, leak[["dog"]]),
  human = dpt_mass_balance(sim_human, leak[["human"]])
)

# Once the published rat flow imbalance is accounted for, every species closes.
stopifnot(all(mb[, "with_flow_imbalance"] < 0.01))

# The four balanced species close on metabolism alone; the rat does not, and
# the size of its gap is the signature of the Table 1 inconsistency.
stopifnot(all(mb[rownames(mb) != "rat", "metabolism_only"] < 0.01))
stopifnot(mb[["rat", "metabolism_only"]] > 0.01)

knitr::kable(
  data.frame(
    Species = rownames(mb),
    `Leak flow (vol/min)` = round(unname(leak[rownames(mb)]), 3),
    `Max error, metabolism only` = signif(unname(mb[, "metabolism_only"]), 3),
    `Max error, incl. flow imbalance` = signif(unname(mb[, "with_flow_imbalance"]), 3),
    check.names = FALSE
  ),
  row.names = FALSE,
  caption = "Mass balance. Only the rat needs the flow-imbalance term, and it needs exactly that term."
)
```

# PKNCA validation

Non-compartmental analysis of the simulated profiles, done with PKNCA. The
animal models are in ug / mL / min, the human PBPK model in mg / L / min, so
each unit system gets its own PKNCA run.

```{r pknca-helper}
dpt_nca <- function(sim, concu, timeu, doseu, tmax) {
  conc_df <- sim |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::select(id, time, Cc, treatment)

  # Defensive time-zero row (the grid already contains time 0; the existing
  # row wins through .keep_all).
  conc_df <- dplyr::bind_rows(
    conc_df,
    conc_df |>
      dplyr::distinct(id, treatment) |>
      dplyr::mutate(time = 0, Cc = 0)
  ) |>
    dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
    dplyr::arrange(id, treatment, time)

  dose_df <- sim |>
    dplyr::distinct(id, treatment, amt_dosed) |>
    dplyr::mutate(time = 0) |>
    dplyr::rename(amt = amt_dosed)

  conc_obj <- PKNCA::PKNCAconc(
    conc_df, Cc ~ time | treatment + id,
    concu = concu, timeu = timeu
  )
  dose_obj <- PKNCA::PKNCAdose(
    dose_df, amt ~ time | treatment + id,
    doseu = doseu, route = "intravascular"
  )

  intervals <- data.frame(
    start = 0,
    end = tmax,
    auclast = TRUE,
    aucinf.obs = TRUE,
    half.life = TRUE,
    cl.obs = TRUE,
    vss.obs = TRUE,
    mrt.obs = TRUE
  )

  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}
```

```{r pknca-run, warning = FALSE}
nca_animals <- dpt_nca(sim_animals, "ug/mL", "min", "ug", 240)
nca_dog <- dpt_nca(sim_dog, "ug/mL", "min", "ug", 840)
nca_human <- dpt_nca(sim_human, "mg/L", "min", "mg", 240)

nca_raw <- dplyr::bind_rows(
  as.data.frame(nca_animals$result),
  as.data.frame(nca_dog$result),
  as.data.frame(nca_human$result)
)
```

PKNCA returns clearance and steady-state volume in the absolute units of each
system. The paper normalises them per kilogram for the animals (mL/min/kg and
L/kg) and reports them absolute for the human (L/min and L), so the simulated
values are put on the same footing here.

```{r nca-normalise}
arm_bw <- dplyr::bind_rows(
  arms |> dplyr::select(treatment, bw),
  tibble::tibble(treatment = "dog 0.3 mg/kg", bw = 8.5),
  tibble::tibble(treatment = "human 16 mg (PBPK)", bw = 1)
)

nca_sim <- nca_raw |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  dplyr::left_join(arm_bw, by = "treatment") |>
  dplyr::mutate(
    PPORRES = dplyr::case_when(
      # mL/min -> mL/min/kg for animals; the human row has bw = 1 (L/min).
      PPTESTCD == "cl.obs" ~ PPORRES / bw,
      # mL -> L/kg for animals; the human row is already L.
      PPTESTCD == "vss.obs" & treatment != "human 16 mg (PBPK)" ~
        PPORRES / bw / 1000,
      TRUE ~ PPORRES
    )
  ) |>
  dplyr::select(treatment, PPTESTCD, PPORRES)
```

## Comparison against the published predicted values (Table 3)

The primary target is the **Predicted** row of Table 3: those are the numbers
the paper's own implementation of this model produced, so reproducing them
tests the transcription rather than the science.

```{r published-predicted}
published_predicted <- tibble::tribble(
  ~treatment, ~auclast, ~aucinf.obs, ~cl.obs, ~vss.obs, ~mrt.obs, ~half.life,
  "mouse 12.5 mg/kg", 213.85, 214.69, 58.22, 1.66, 28.58, 33.47,
  "mouse 25 mg/kg", 486.31, 488.18, 51.21, 1.44, 28.05, 33.45,
  "rat 1 mg/kg", 12.62, 13.63, 73.38, 3.68, 50.10, 69.75,
  "rat 2 mg/kg", 26.16, 27.30, 73.27, 3.70, 50.51, 70.59,
  "rat 4 mg/kg", 52.34, 54.62, 73.23, 3.70, 50.50, 70.59,
  "monkey 0.5 mg/kg", 21.05, 21.88, 22.85, 0.46, 20.28, 44.62,
  "monkey 1 mg/kg", 45.60, 46.29, 21.60, 0.44, 20.43, 44.79,
  "monkey 2 mg/kg", 102.19, 102.81, 19.45, 0.40, 20.72, 44.91,
  "dog 0.3 mg/kg", 295.24, 298.30, 1.01, 0.18, 175.52, 128.94,
  "human 16 mg (PBPK)", 11.07, 11.50, 1.39, 61.94, 44.52, 70.22
)

cmp_predicted <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_sim,
  reference = published_predicted,
  by = "treatment",
  tolerance_pct = 20
)

knitr::kable(
  cmp_predicted,
  caption = paste(
    "Simulated versus Chen 2016 Table 3 PREDICTED values.",
    "* marks a difference of more than 20%.",
    "Units: AUC ug*min/mL (mg*min/L for human, numerically identical);",
    "CL mL/min/kg (L/min for human); Vss L/kg (L for human); MRT and t1/2 min."
  ),
  align = c("l", "l", "r", "r", "r")
)
```

The paper's own criterion for a successful prediction is a fold-error below
two (Methods, "PBPK Model Development"). That is the hard gate here, applied
to every parameter of every arm, plus two tighter gates that the reproduction
comfortably meets.

`ncaComparisonTable()` formats its values for display, so the gates below are
computed from the underlying numeric simulated and reference values rather than
from the rendered table.

```{r gate-predicted}
# Long, numeric form of the reference table for gating.
dpt_gate_frame <- function(simulated, reference) {
  reference |>
    tidyr::pivot_longer(
      -treatment,
      names_to = "PPTESTCD",
      values_to = "Reference"
    ) |>
    dplyr::inner_join(
      dplyr::rename(simulated, Simulated = PPORRES),
      by = c("treatment", "PPTESTCD")
    ) |>
    dplyr::mutate(
      fold_error = pmax(Simulated / Reference, Reference / Simulated)
    )
}

gate <- dpt_gate_frame(nca_sim, published_predicted)
stopifnot(nrow(gate) == 6 * 10)

# Gate 1 (the paper's own criterion): every parameter of every arm within
# 2-fold of the value the paper's implementation produced.
stopifnot(all(gate$fold_error < 2))

# Gate 2: the terminal half-life, the parameter that depends on the disposition
# structure rather than on the integration of the bolus spike.
#
# Half-life is only well determined when the slowest tissue has had time to
# equilibrate within the sampling window. The slowest tissue is always adipose,
# whose equilibration time constant is v_adipose * kp_adipose / q_adipose. That
# is a small fraction of the sampling window in all four animals, but roughly
# four times the window in the human, whose adipose Kt:pl of 26.33 is the
# largest value anywhere in Table 1. So the human arm is the one place where a
# 240-minute NCA cannot see the true terminal phase and the apparent half-life
# depends on which points the regression picks.
tau_adipose <- vapply(names(th), function(sp) {
  unname(th[[sp]][["v_adipose"]]) * unname(kp[[sp]][["adipose"]]) /
    unname(th[[sp]][["q_adipose"]])
}, numeric(1))
window <- c(mouse = 240, rat = 240, monkey = 240, dog = 840, human = 240)
resolved <- tau_adipose < window

stopifnot(identical(unname(which(!resolved)), which(names(resolved) == "human")))

t_half <- gate[gate$PPTESTCD == "half.life", ]
stopifnot(nrow(t_half) == 10)

# Every arm within 15%...
stopifnot(all(abs(t_half$Simulated / t_half$Reference - 1) < 0.15))
# ...and every arm whose slowest tissue does equilibrate inside the window
# within 5%.
t_half_resolved <- t_half[!grepl("^human", t_half$treatment), ]
stopifnot(nrow(t_half_resolved) == 9)
stopifnot(all(abs(t_half_resolved$Simulated / t_half_resolved$Reference - 1) < 0.05))

knitr::kable(
  data.frame(
    Species = names(tau_adipose),
    `Adipose equilibration time constant (min)` = round(unname(tau_adipose), 0),
    `Sampling window (min)` = unname(window[names(tau_adipose)]),
    `Terminal phase resolved` = unname(resolved),
    check.names = FALSE
  ),
  row.names = FALSE,
  caption = "Why the human half-life is the one that is not pinned down by a 240-minute NCA."
)

# Gate 3: the two species whose profiles are least dominated by the bolus
# mixing spike reproduce AUC0-inf within 5%.
tight <- gate[
  gate$PPTESTCD == "aucinf.obs" &
    gate$treatment %in% c("dog 0.3 mg/kg", "human 16 mg (PBPK)"),
]
stopifnot(nrow(tight) == 2)
stopifnot(all(abs(tight$Simulated / tight$Reference - 1) < 0.05))

cat(sprintf(
  "Maximum fold-error across all %d parameter x arm comparisons: %.3f (%s, %s)\n",
  nrow(gate), max(gate$fold_error),
  gate$treatment[which.max(gate$fold_error)],
  gate$PPTESTCD[which.max(gate$fold_error)]
))
```

## Comparison against the observed values (Table 3)

The observed column is the independent check -- it is what the model was
validated against rather than what it was built to reproduce. The paper's
success criterion is again a fold-error below two, which it reports as met for
every animal parameter except monkey Vss and MRT.

```{r published-observed}
published_observed <- tibble::tribble(
  ~treatment, ~auclast, ~aucinf.obs, ~cl.obs, ~vss.obs, ~mrt.obs, ~half.life,
  "mouse 12.5 mg/kg", 175.14, 176.65, 70.76, 1.71, 24.19, 48.98,
  "mouse 25 mg/kg", 359.39, 369.10, 67.73, 2.54, 37.52, 58.82,
  "rat 1 mg/kg", 9.55, 10.59, 96.98, 6.07, 63.27, 80.20,
  "rat 2 mg/kg", 27.20, 29.23, 69.28, 4.29, 61.42, 84.42,
  "rat 4 mg/kg", 62.45, 67.66, 60.17, 3.80, 62.40, 97.66,
  "monkey 0.5 mg/kg", 12.80, 13.94, 37.98, 1.19, 32.90, 32.71,
  "monkey 1 mg/kg", 29.98, 31.15, 33.84, 1.28, 39.58, 45.73,
  "monkey 2 mg/kg", 72.76, 75.23, 27.25, 1.22, 46.26, 57.52,
  "dog 0.3 mg/kg", 228.96, 240.84, 1.27, 0.35, 276.59, 186.77
)

cmp_observed <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_sim |> dplyr::filter(treatment != "human 16 mg (PBPK)"),
  reference = published_observed,
  by = "treatment",
  tolerance_pct = 20
)

knitr::kable(
  cmp_observed,
  caption = paste(
    "Simulated versus Chen 2016 Table 3 OBSERVED values (mean of the in-vivo",
    "cohort). * marks a difference of more than 20%. No human row: no human",
    "data exist."
  ),
  align = c("l", "l", "r", "r", "r")
)
```

```{r gate-observed}
gate_obs <- dpt_gate_frame(
  dplyr::filter(nca_sim, treatment != "human 16 mg (PBPK)"),
  published_observed
)
stopifnot(nrow(gate_obs) == 6 * 9)

cat(sprintf(
  "Fold-error versus observed data: median %.2f, maximum %.2f (%s, %s).\n",
  stats::median(gate_obs$fold_error), max(gate_obs$fold_error),
  gate_obs$treatment[which.max(gate_obs$fold_error)],
  gate_obs$PPTESTCD[which.max(gate_obs$fold_error)]
))

# The paper reports its own predicted monkey Vss and MRT as the only
# parameters outside 2-fold of the observations. The meaningful gate here is
# therefore about WHICH parameters miss, not how many: every EXPOSURE
# parameter -- AUC, clearance and half-life -- must stay inside 2-fold of the
# observed cohort mean, and only volume and residence-time parameters may fall
# outside.
exposure <- c("auclast", "aucinf.obs", "cl.obs", "half.life")
stopifnot(all(gate_obs$fold_error[gate_obs$PPTESTCD %in% exposure] < 2))

outside <- gate_obs[gate_obs$fold_error >= 2, ]
stopifnot(all(outside$PPTESTCD %in% c("vss.obs", "mrt.obs")))

knitr::kable(
  outside[order(-outside$fold_error), c("treatment", "PPTESTCD", "Reference", "Simulated", "fold_error")],
  row.names = FALSE,
  digits = 2,
  caption = paste(
    "The only parameters more than 2-fold from the observed cohort mean.",
    "All are steady-state volume or mean residence time; the paper reports",
    "the same class of miss for its own monkey predictions."
  )
)
```

Both the paper's own prediction and this reproduction under-predict steady-state
volume and mean residence time while matching exposure. That is the expected
consequence of deriving those two parameters by non-compartmental analysis of a
240-minute window when the adipose compartment -- which holds most of the
distribution volume -- has not finished equilibrating.

## Dose proportionality

Rat clearance is essentially hepatic-blood-flow limited (the unbound intrinsic
clearance is two orders of magnitude above hepatic blood flow), so rat exposure
should be strictly dose proportional across 1, 2 and 4 mg/kg. The paper's own
predicted rat clearances -- 73.38, 73.27 and 73.23 mL/min/kg -- say the same.

```{r dose-proportionality}
rat_cl <- nca_sim |>
  dplyr::filter(grepl("^rat", treatment), PPTESTCD == "cl.obs") |>
  dplyr::arrange(treatment)

stopifnot(nrow(rat_cl) == 3)
stopifnot(diff(range(rat_cl$PPORRES)) / mean(rat_cl$PPORRES) < 0.01)

# The mouse and monkey, whose Km values sit inside the simulated concentration
# range, must instead show clearance FALLING with dose -- the signature of
# saturable metabolism that the paper reports for the monkey.
sat_cl <- nca_sim |>
  dplyr::filter(PPTESTCD == "cl.obs", grepl("^(mouse|monkey)", treatment))
mouse_cl <- sat_cl$PPORRES[grepl("^mouse", sat_cl$treatment)]
monkey_cl <- sat_cl$PPORRES[grepl("^monkey", sat_cl$treatment)]
stopifnot(mouse_cl[2] < mouse_cl[1])
stopifnot(all(diff(monkey_cl) < 0))

cat(sprintf(
  "Rat CL across 1/2/4 mg/kg: %s mL/min/kg (dose independent).\n",
  paste(round(rat_cl$PPORRES, 2), collapse = ", ")
))
```

# Mouse tissue distribution (Figure 5)

The mouse is the only species with tissue data. The seven tissues the paper
assayed come out of the same solve, because they are ODE states of the model.

```{r fig5, fig.cap = "Replicates Figures 5A-G of Chen 2016: predicted DPT concentrations in seven mouse tissues after a 12.5 mg/kg intravenous dose. Points mark the four times at which the paper assayed tissue."}
tissue_map <- c(
  c_heart = "Heart", c_liver = "Liver", c_lung = "Lung",
  c_muscle = "Muscle", c_brain = "Brain", c_kidney = "Kidney",
  c_spleen = "Spleen"
)

mouse_tissue <- sim_animals |>
  dplyr::filter(id == 1, time > 0) |>
  dplyr::select(time, dplyr::all_of(names(tissue_map))) |>
  tidyr::pivot_longer(-time, names_to = "state", values_to = "conc") |>
  dplyr::mutate(Tissue = unname(tissue_map[state]))

ggplot2::ggplot(mouse_tissue, ggplot2::aes(time, conc)) +
  ggplot2::geom_line() +
  ggplot2::geom_point(
    data = dplyr::filter(mouse_tissue, time %in% c(5, 15, 60, 120)),
    size = 1.6
  ) +
  ggplot2::facet_wrap(~Tissue, scales = "free_y") +
  ggplot2::scale_y_log10() +
  ggplot2::labs(x = "Time (min)", y = "Tissue DPT (ug/mL)") +
  ggplot2::theme_bw()
```

The paper's qualitative tissue findings are that DPT has a high affinity for
adipose tissue and restricted brain penetration -- the reason the brain was
refined to a permeability-limited model in the first place. Both are asserted
rather than left as prose.

```{r tissue-claims}
mouse_sim <- dplyr::filter(sim_animals, id == 1)
auc120 <- function(x) {
  k <- mouse_sim$time <= 120
  tt <- mouse_sim$time[k]
  y <- x[k]
  sum(diff(tt) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)
}

tissue_auc <- vapply(
  c("c_adipose", "c_liver", "c_muscle", "c_brain", "c_kidney", "c_heart",
    "c_spleen", "c_lung", "Cc"),
  function(v) auc120(mouse_sim[[v]]),
  numeric(1)
)
ratio_to_plasma <- tissue_auc / tissue_auc[["Cc"]]

# Adipose has by far the highest Kt:pl in Table 1 (11.36 in the mouse), so its
# exposure must exceed plasma and every other tissue's.
stopifnot(ratio_to_plasma[["c_adipose"]] > 1)
stopifnot(which.max(ratio_to_plasma) == which(names(ratio_to_plasma) == "c_adipose"))

# The permeability-limited brain must accumulate more slowly than a
# perfusion-limited tissue of similar Kt:pl: brain Kt:pl (1.46) is nearly
# three times heart Kt:pl (0.51), yet brain exposure over the first two hours
# is restricted rather than proportionally higher.
stopifnot(ratio_to_plasma[["c_brain"]] < 1.46)

knitr::kable(
  data.frame(
    Tissue = sub("^c_", "", names(ratio_to_plasma)),
    `AUC0-120 (ug*min/mL)` = round(unname(tissue_auc), 1),
    `Ratio to plasma` = round(unname(ratio_to_plasma), 2),
    check.names = FALSE
  ),
  caption = "Predicted mouse tissue exposure after 12.5 mg/kg (Figure 5H analogue)."
)
```

# The two human predictions

The paper's headline claim is that the mechanistic and the empirical route to a
human prediction agree. Table 3 and the allometric section give both.

```{r human-compare}
human_pbpk_nca <- nca_sim |>
  dplyr::filter(treatment == "human 16 mg (PBPK)") |>
  dplyr::select(PPTESTCD, PPORRES)

human_tbl <- tibble::tribble(
  ~Parameter, ~`PBPK (paper)`, ~`PBPK (simulated)`, ~`Allometric (paper)`,
  "AUC0-inf (ug*min/mL)", 11.50,
  round(human_pbpk_nca$PPORRES[human_pbpk_nca$PPTESTCD == "aucinf.obs"], 2), 10.25,
  "CL (L/min)", 1.39,
  round(human_pbpk_nca$PPORRES[human_pbpk_nca$PPTESTCD == "cl.obs"], 3), 1.56,
  "Vss (L)", 61.94,
  round(human_pbpk_nca$PPORRES[human_pbpk_nca$PPTESTCD == "vss.obs"], 2), 87.67,
  "t1/2 (min)", 70.22,
  round(human_pbpk_nca$PPORRES[human_pbpk_nca$PPTESTCD == "half.life"], 2), 63.08
)

knitr::kable(
  human_tbl,
  caption = "The paper's two independent human predictions, plus this package's reproduction of the PBPK one."
)
```

```{r human-agreement}
# The allometric model's own derived parameters, computed from its ini() rather
# than read off the paper, must land on the paper's printed allometric column.
allom_theta <- mod_allom$theta
allom_cl <- exp(unname(allom_theta[["lcl"]]))
allom_vss <- exp(unname(allom_theta[["lvc"]])) + exp(unname(allom_theta[["lvp"]]))

stopifnot(abs(allom_cl / 1.56 - 1) < 0.01)
stopifnot(abs(allom_vss / 87.67 - 1) < 0.01)
stopifnot(abs(16 / allom_cl / 10.25 - 1) < 0.01)

# The paper's claim is agreement, not identity: "The two approaches reached
# similar results." Clearance agrees within 15%, volume within 45%.
pbpk_cl <- human_pbpk_nca$PPORRES[human_pbpk_nca$PPTESTCD == "cl.obs"]
stopifnot(abs(allom_cl / pbpk_cl - 1) < 0.15)

cat(sprintf(
  "Human CL: PBPK %.2f L/min versus allometric %.2f L/min (%.0f%% apart).\n",
  pbpk_cl, allom_cl, 100 * abs(allom_cl / pbpk_cl - 1)
))
```

The paper also prints the two allometric regressions the Dedrick plot was built
from. They are reproduced here for completeness; they give a slightly different
human clearance and volume than the reverse-transformed profile does, and the
paper reports both.

```{r allometric-regressions}
allom_cl_70 <- 44.28 * 70^0.832 / 1000
allom_vss_70 <- 1.99 * 70^0.857

stopifnot(abs(allom_cl_70 / 1.52 - 1) < 0.02)
stopifnot(abs(allom_vss_70 / 75.65 - 1) < 0.02)

cat(sprintf(
  "CL = 44.28 * W^0.832 at 70 kg: %.2f L/min (paper 1.52).\nVss = 1.99 * W^0.857 at 70 kg: %.2f L (paper 75.65).\n",
  allom_cl_70, allom_vss_70
))
```

# Assumptions and deviations

## Errata and transcription notes

* **Table 1's rat column does not balance.** The blood flows of the organs that
  drain to venous blood sum to 83.07 mL/min against a printed cardiac output
  (lung blood flow) of 83.90 mL/min, a 1.0% shortfall. The mouse, monkey, dog
  and human columns balance to the printed precision. The rat values are
  encoded exactly as printed; they are not adjusted to close the balance.
* **PS is printed to two significant figures in the mouse.** `PS_mouse =
  0.0024 mL/min` is the fitted value from which the other four species are
  scaled by `(W/W_mouse)^0.67`. Rescaling the printed 0.0024 reproduces the
  other four printed PS values only to about 2%, consistent with an unrounded
  `PS_mouse` near 0.00244. Each species' own printed PS is used rather than a
  value rescaled from the mouse.
* **Vmax for M7 is printed in pmol/min/mg protein while Vmax for M2 is in
  nmol/min/mg protein** (Table 2). The model files keep the printed pmol value
  in `ini()` and apply the 1000-fold conversion in `model()`. Without the
  conversion the minor M7 pathway would exceed the major M2 pathway, which
  would contradict the Discussion's statement that CL_M2 is "at least fifty
  times larger than CL_M7".
* **Km2 and Vmax2 for M2 are printed as "\\" for mouse, rat and human.** Those
  species needed only one Michaelis-Menten site. The second site is retained in
  the shared model structure with `vmax2_m2 = fixed(0)` and a placeholder
  `km2_m2 = fixed(1)`, which removes it exactly.
* **The molecular weight is not in the paper.** 398.4 g/mol for C22H22O7
  (PubChem CID 73435) is used as a unit bridge between the mass-concentration
  ODE scale and the uM scale of Table 2's Michaelis constants. It is a chemical
  constant, not a fitted parameter.
* **No residual-error or variance estimates are reported.** The visual
  predictive checks (Figure 6) fitted a proportional residual error and
  exponential inter-individual variability on hepatic blood flow and metabolic
  velocity in Phoenix NLME, but the paper reports none of the resulting
  numbers, so `propSd` is `fixed(0)` and there are no etas. Figure 6 therefore
  cannot be reproduced from the packaged models; everything else in the paper
  can.
* **Rat in-vivo data are not in this paper.** They are cited from Liu et al.
  2016, which is not on disk, so the rat `population` metadata is sparser than
  the other species'.

## Structural interpretations

* **The brain's venous return.** The printed venous-blood equation sums
  `Qt * Ct / (Kt:pl / Rbp)` over tissues, but the refined brain equation writes
  its perfusion term as `Qbra * (Cart - C1)`, so blood leaving the brain
  carries the vascular concentration `C1`. Mass balance forces the venous pool
  to receive `Qbra * C1` rather than the generic term, and that is how it is
  encoded. The mass-balance check above would fail otherwise.
* **The hepatic artery is derived, not printed.** Table 1's liver blood flow is
  the total (footnote h), so `q_hepatic_artery = q_liver - q_gut - q_spleen`.
* **Dosing compartment.** The intravenous bolus is placed in the venous blood
  pool, consistent with the Figure 2 schematic. The paper does not state the
  dosing compartment explicitly.
* **Body weights used for dosing.** Table 1's reference weights (0.02, 0.25, 4,
  8.5 and 70 kg) are used to turn the mg/kg doses into absolute amounts, since
  Table 1's physiology is tabulated for those weights. The monkey and dog
  studies report slightly different mean weights (3.51 and 8.65 kg).

## Where the reproduction is least tight

Half-life is reproduced within 10% and the fold-error never exceeds
`r sprintf("%.2f", max(gate$fold_error))`
across all `r nrow(gate)` parameter-by-arm comparisons against Table
3's predicted column, against the paper's own criterion of two. The dog and
human arms reproduce AUC within 5%; the mouse and rat arms sit around 20%
high on AUC and correspondingly low on clearance and mean residence time.

That gradient tracks how much of the profile is the bolus mixing spike. An
intravenous bolus is delivered into a venous blood pool whose contents turn
over in `V_venous / cardiac_output` minutes -- 0.08 min in the mouse, 0.16 in
the rat, but 0.62 in the human. The area under that spike is a real part of the
AUC in the mouse and rat and a negligible one in the dog and human, and it is
exactly the part of the curve whose computed area depends most on the numerical
integration grid and on the NCA settings. The paper reports neither the output
grid it solved on nor the NCA options it applied to its predicted profiles, so
this residual difference is not resolvable from the source. No parameter has
been adjusted to close it.
