---
title: "Biapenem (Chen 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Biapenem (Chen 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Chen_2024_biapenem"))$reference`
* Description: `r rxode2::rxode(readModelDb("Chen_2024_biapenem"))$description`
* Article: <https://doi.org/10.3389/fphar.2024.1388150>
* Supplement (Table S1 covariate screening, Figure S1 concentration scatter):
  <https://www.frontiersin.org/articles/10.3389/fphar.2024.1388150/full#supplementary-material>

Biapenem is a carbapenem used widely in Asia. Chen 2024 is the first population
PK analysis of biapenem in septic patients, and its practical conclusion is that
the product-label maximum of 1.2 g/day is too low for this population.

```{r model}
mod <- rxode2::rxode(readModelDb("Chen_2024_biapenem"))
```

# Population

The model was fit to a **modeling cohort of 245 adults with sepsis** (351
biapenem concentrations) drawn from 317 patients treated at Nanjing Drum Tower
Hospital, China, between January 2018 and May 2022; the remaining 72 patients
(115 concentrations) formed an external evaluation cohort. Sepsis was defined by
the Sepsis-3 criteria. Median age was 63 years (range 18-97), median weight 62 kg
(36.8-100), and 36.3% were female. Septic shock was present in 26.9% and 13.9%
were immunocompromised; the primary infection site was intra-abdominal in 48.6%,
respiratory in 44.1%.

Renal function spanned nearly the full clinical range: Cockcroft-Gault
creatinine clearance median 84.92 mL/min (range 3.5-295.5) and blood urea
nitrogen median 6.2 mmol/L (range 0.4-66.9). Patients on renal replacement
therapy or ECMO were excluded.

Sampling was sparse and opportunistic -- routine therapeutic drug monitoring,
averaging **1.5 samples per participant**, usually drawn after the third dosing
interval, with a median time after the last dose of 6 h. This matters for
interpreting the model: with roughly one sample per patient there is very little
information with which to separate between-subject variability from residual
error, and the estimated variance components should be read with that in mind
(see [Errata](#errata-and-adjudications)).

Baseline demographics are Chen 2024 Table 1 (modeling-cohort columns). The same
information is available programmatically:

```{r population}
str(mod$population)
```

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (theta1) | 8.33 L/h | Table 2, final-model estimate (RSE 6.4%) |
| `lvc` (theta2) | 13.4 L | Table 2, final-model estimate (RSE 15%) |
| `lq` (theta3) | 3.75 L/h | Table 2, final-model estimate (RSE 13.6%) |
| `lvp` (theta4) | 60.4 L | Table 2, final-model estimate (RSE 17.5%) |
| `e_crcl_cl` (theta5) | 0.0046 per mL/min | Table 2 (RSE 11.9%); see Errata re. the in-text `0.046` |
| `e_bun_q` (theta6) | 0.112 per mmol/L | Table 2 (RSE 1.9% as printed) |
| CRCL centering | 78.2 mL/min | Table 2 CL equation; Discussion ("8.33 L/h for CLCr of 78.2 mL/min") |
| BUN centering | 6.8 mmol/L | Table 2 Q equation |
| `etalcl` | 0.0591 | Table 2, inter-individual variability (RSE 19.8%) |
| `etalq` | 1.12 | Table 2, inter-individual variability (RSE 25.2%) |
| `addSd` | sqrt(0.591) = 0.769 mg/L | Table 2, residual variability (RSE 17.1%) |
| Exponential IIV | `P_i = TV(P) * exp(eta_i)` | Equation 1 |
| Additive residual error | `Y = IPRED + eps` | Equation 2; Results 3.2 selects the additive form |
| Two-compartment first-order ODEs | n/a | Results 3.2 (OFV 679.233 two-cmt vs 734.914 one-cmt) |
| Covariate screening / OFV chain | n/a | Supplementary Table S1 (base 683.499 -> final 550.508) |

```{r source-trace-check}
# The published structural values must be exactly what is encoded.
theta <- setNames(mod$theta, names(mod$theta))
stopifnot(
  isTRUE(all.equal(unname(exp(theta[["lcl"]])), 8.33, tolerance = 1e-8)),
  isTRUE(all.equal(unname(exp(theta[["lvc"]])), 13.4, tolerance = 1e-8)),
  isTRUE(all.equal(unname(exp(theta[["lq"]])),   3.75, tolerance = 1e-8)),
  isTRUE(all.equal(unname(exp(theta[["lvp"]])), 60.4, tolerance = 1e-8)),
  isTRUE(all.equal(unname(theta[["e_crcl_cl"]]), 0.0046, tolerance = 1e-12)),
  isTRUE(all.equal(unname(theta[["e_bun_q"]]),   0.112,  tolerance = 1e-12))
)

# Chen 2024 Discussion states V(V1+V2) = 73.8 L for this cohort, "approximately
# four to five times higher than that observed in other patient populations".
v_total <- exp(theta[["lvc"]]) + exp(theta[["lvp"]])
stopifnot(isTRUE(all.equal(unname(v_total), 73.8, tolerance = 1e-8)))
c(V1_plus_V2 = unname(v_total))
```

# Errata and adjudications {#errata-and-adjudications}

Three points in the paper required adjudication before the values above could be
encoded. Each is recorded here with the evidence, because two of them change a
number and the third is unresolved.

## 1. The in-text CL equation has lost a zero (resolved)

Results 3.2 prints the final model as

> `CL (L/h) = 8.33 x [1 + 0.046 x (CLCr (mL/min) - 78.2)]`

but Table 2 gives `theta5 = 0.0046` -- a factor of ten smaller. **Table 2 is
correct.** The in-text value is falsified by the paper's own data in two
independent ways:

```{r theta5-falsification}
crcl_grid <- c(3.5, 25, 56.5, 78.2, 150, 295.5)   # spans the observed range
cl_table <- 8.33 * (1 + 0.0046 * (crcl_grid - 78.2))   # Table 2
cl_text  <- 8.33 * (1 + 0.046  * (crcl_grid - 78.2))   # in-text equation

data.frame(CRCL = crcl_grid, CL_Table2 = cl_table, CL_intext = cl_text) |>
  dplyr::rename("CLCr (mL/min)" = CRCL,
                "CL, Table 2 (L/h)" = CL_Table2,
                "CL, in-text eq. (L/h)" = CL_intext) |>
  knitr::kable(digits = 2, caption = paste(
    "The in-text coefficient 0.046 drives clearance negative across much of the",
    "observed CLCr range; the Table 2 coefficient 0.0046 does not."
  ))

# (a) 0.046 makes CL negative below CLCr = 78.2 - 1/0.046 = 56.5 mL/min, which
#     covers a large share of a cohort whose CLCr median is 84.92 and whose
#     range starts at 3.5. 0.0046 would require CLCr = -139 mL/min, unreachable.
zero_at_text  <- 78.2 - 1 / 0.046
zero_at_table <- 78.2 - 1 / 0.0046
stopifnot(zero_at_text > 3.5, zero_at_table < 0)

# (b) The bootstrap 5th-95th percentile interval for theta5 is 0.0034-0.0064,
#     which brackets 0.0046 and excludes 0.046 by two orders of magnitude.
stopifnot(0.0034 < 0.0046, 0.0046 < 0.0064, 0.046 > 0.0064)
c(CL_zero_crossing_intext = zero_at_text, CL_zero_crossing_Table2 = zero_at_table)
```

## 2. The variance components are read as variances, not standard deviations (unresolved; assumption stated)

Table 2 labels its variability rows `omega CL`, `omega Q` and `eps`, while
Methods 2.3.1 defines `eta_i` as having "variance of `omega^2`" and `eps` as
having "variance of `sigma^2`". Read literally against that notation the
tabulated numbers would be *standard deviations*; read as NONMEM output -- which
prints the `$OMEGA` and `$SIGMA` diagonal elements, i.e. variances -- they are
*variances*. The paper contains no control stream and the supplement (Table S1,
Figure S1) does not resolve it.

**This model encodes them as variances.** The reasoning:

* Read as a standard deviation, `omega CL = 0.0591` implies a **5.9% coefficient
  of variation on clearance**. That directly contradicts the paper's own
  motivation -- the Introduction argues biapenem "manifests a significant
  inter-individual variability in its pharmacokinetic characteristics, which
  leads to substantial disparities in the blood drug concentration amongst
  patients", which is why TDM and dose individualisation are needed at all.
  Read as a variance it implies 24.7% CV, the ordinary magnitude for a renally
  cleared antibiotic after creatinine clearance is already in the model.
* The bootstrap 5th-95th interval for `omega CL` is 0.031-0.088. As variances
  that is a CL-IIV range of CV 17.6-29.7%, which a sparse dataset can plausibly
  deliver. As standard deviations it is CV 3.1-8.8% -- asserting that 351
  samples from 245 patients pinned clearance variability to single digits, which
  1.4 samples per patient cannot support.
* The bootstrap interval for `omega Q` (0.26 / 1.10 / 2.94) is strongly
  right-skewed, the signature of a variance estimate rather than a
  standard-deviation estimate.

```{r omega-scale}
cv <- function(x) 100 * sqrt(exp(x^2) - 1)   # log-normal CV% from a log-scale SD
data.frame(
  Parameter = c("omega CL", "omega CL", "omega Q", "omega Q"),
  Reading   = c("variance (encoded)", "standard deviation",
                "variance (encoded)", "standard deviation"),
  log_sd    = c(sqrt(0.0591), 0.0591, sqrt(1.12), 1.12),
  CV_pct    = c(cv(sqrt(0.0591)), cv(0.0591), cv(sqrt(1.12)), cv(1.12))
) |>
  dplyr::rename("Table 2 row" = Parameter, "Interpretation" = Reading,
                "log-scale SD" = log_sd, "CV (%)" = CV_pct) |>
  knitr::kable(digits = c(0, 0, 3, 1), caption = paste(
    "The two readings differ ~4-fold on CL but are nearly identical on Q,",
    "so only the clearance term is materially affected."
  ))
```

Note that the two readings differ very little for `omega Q` (log-scale SD 1.06
vs 1.12), so the *gross* behaviour of the model -- which is dominated by that
very large distributional variability -- is similar either way. The choice
matters for clearance, and therefore for exposure and target attainment.

## 3. The published Monte Carlo target attainment is not reproducible from the published parameters (documented deviation)

Chen 2024 Figure 3 reports probability of attaining 70% *f*T>MIC by regimen and
MIC. Those curves cannot be reproduced from Table 2 under **either** reading of
the variance components, and the reason is visible in the figure itself: at
MIC = 0.0625 mg/L the reported attainment plateaus at only ~70% (300 mg q12h) to
~93% (600 mg q6h). With a typical trough of about 2 mg/L and a terminal
half-life near 16 h, essentially every virtual patient should hold
concentrations above 0.0625 mg/L for the whole interval, giving ~100%. A
plateau well below 100% at a negligible MIC requires a large subpopulation with
near-zero exposure, which this parameter table does not generate.

The [target attainment](#target-attainment) section below reproduces the
analysis and quantifies the gap rather than tuning to it. **No parameter was
adjusted to match the published curves.**

## Other notes

* Serum albumin reached significance on CL during forward inclusion
  (`dOFV = -3.891`) but was removed in backward elimination
  (`dOFV = +0.285 < 6.63`) and is not in the final model. No coefficient is
  reported for it anywhere, so it cannot be encoded; it is recorded in the model
  file under `covariatesDataExcluded`.
* `theta6` is printed with an RSE of 1.9%, which is inconsistent with its own
  bootstrap interval (0.034-0.15, a four-fold spread). This looks like a typo
  for a larger RSE, but it does not affect the point estimate and no value was
  changed.
* Biapenem's unbound fraction is not reported. *f*T>MIC below therefore uses
  total plasma concentration; biapenem is only lightly protein bound, so the
  correction would be small, but it is an assumption, not a published value.

# Virtual cohort

The original data are not public. The cohort below reproduces the modeling-cohort
covariate distributions of Table 1 -- both are strongly right-skewed, so each is
drawn log-normally from its published median and mean and truncated to its
published range.

```{r cohort}
# set.seed() seeds R's RNG only. rxode2's simulation RNG is partitioned per
# solver thread, so a machine with a different thread count draws a different
# cohort. Every assertion below is written to hold for any cohort this model
# can produce (see known-vignette-failure-patterns.md pattern 12).
set.seed(20240510)

n_per_arm <- 200L   # cap is 200/arm

# Log-normal matched to the published median and mean, truncated to the range.
skewed <- function(n, median_, mean_, lo, hi) {
  sdlog <- sqrt(2 * log(mean_ / median_))
  pmin(pmax(rlnorm(n, meanlog = log(median_), sdlog = sdlog), lo), hi)
}

regimens <- data.frame(
  regimen = c("300 mg q6h", "300 mg q8h", "600 mg q6h", "600 mg q8h"),
  amt     = c(300, 300, 600, 600),
  tau     = c(6, 8, 6, 8)
)

# COMMON RANDOM NUMBERS: one covariate/eta draw reused across every regimen, so
# regimens differ only by dose and interval. rxSolve() redraws etas on every
# call, so the etas are drawn here in base R and carried as data columns.
subj <- data.frame(
  subject = seq_len(n_per_arm),
  CRCL    = skewed(n_per_arm, 84.92, 88.91, 3.5, 295.5),
  BUN     = skewed(n_per_arm,  6.20,  9.84, 0.4, 66.9),
  etalcl  = rnorm(n_per_arm, 0, sqrt(0.0591)),
  etalq   = rnorm(n_per_arm, 0, sqrt(1.12))
)

summary(subj[, c("CRCL", "BUN")])
```

```{r events}
# Steady-state dosing (ss = 1) gives the exact steady state without a burn-in,
# and one full dosing interval of dense observations.
#
# Observation rows point at "central" -- the ODE STATE -- never at the
# algebraic observable Cc. Naming an observable as a compartment injects a slot
# after the ODE states and renumbers them
# (known-vignette-failure-patterns.md pattern 2). rxode2 returns Cc as an output
# column regardless of which compartment the observation row points at.
build_arm <- function(i) {
  r   <- regimens[i, ]
  ev  <- rxode2::et(amt = r$amt, dur = 1, ii = r$tau, ss = 1, cmt = "central")
  ev  <- rxode2::et(ev, seq(0, r$tau, length.out = 241), cmt = "central")
  ev  <- as.data.frame(ev)
  m   <- nrow(ev)
  out <- ev[rep(seq_len(m), n_per_arm), , drop = FALSE]
  out$id      <- rep(seq_len(n_per_arm), each = m) + (i - 1L) * n_per_arm
  out$subject <- rep(subj$subject, each = m)
  out <- cbind(out, subj[match(out$subject, subj$subject),
                         c("CRCL", "BUN", "etalcl", "etalq")])
  out$regimen <- r$regimen
  out$amt_dose <- r$amt
  out$tau      <- r$tau
  out
}

events <- do.call(rbind, lapply(seq_len(nrow(regimens)), build_arm))
events <- events[order(events$id, events$time, -events$evid), ]
rownames(events) <- NULL
nrow(events)
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod, events,
  omega = NA,                       # etas supplied as data columns
  # Only what is actually read back from `sim`. `tau` and `amt_dose` are read
  # from `regimens` / `events` instead, so they are not carried through the
  # solve -- fewer chances of colliding with an rxode2 name.
  keep  = c("regimen", "subject"),
  returnType = "data.frame"
)
stopifnot(nrow(sim) > 0, !all(is.na(sim$Cc)), all(sim$Cc >= 0))
```

```{r profiles, fig.width = 7, fig.height = 4.5}
sim |>
  group_by(regimen, time) |>
  summarise(
    p05 = quantile(Cc, 0.05), p50 = median(Cc), p95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.2, fill = "steelblue") +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  facet_wrap(~regimen) +
  scale_y_log10() +
  labs(
    x = "Time within the steady-state dosing interval (h)",
    y = "Biapenem concentration (mg/L)",
    title = "Steady-state profiles by regimen",
    caption = paste(
      "Median with 5th-95th percentile band, 200 virtual patients per arm.",
      "Dashed line = 1 mg/L, the MIC at which Chen 2024 evaluates attainment."
    )
  )
```

# PKNCA validation

Steady-state NCA over the dosing interval, run once per regimen because the
regimens have different intervals.

```{r pknca}
# Filter on !is.na(Cc) only. Adding `time > 0` or `Cc > 0` would drop the
# interval-start row that PKNCA needs to anchor the AUC.
run_nca <- function(reg) {
  tau_r <- regimens$tau[regimens$regimen == reg]

  sim_nca <- sim |>
    filter(regimen == reg, !is.na(Cc)) |>
    select(id, time, Cc, regimen)

  dose_df <- events |>
    filter(regimen == reg, evid != 0, time == 0) |>
    select(id, time, amt, regimen) |>
    distinct()

  conc_obj <- PKNCA::PKNCAconc(
    sim_nca, Cc ~ time | regimen + id, concu = "mg/L", timeu = "h"
  )
  dose_obj <- PKNCA::PKNCAdose(
    dose_df, amt ~ time | regimen + id, doseu = "mg"
  )

  # Two-compartment IV infusion declines monotonically after the end of
  # infusion, so over the steady-state interval the minimum IS the trough.
  intervals <- data.frame(
    start = 0, end = tau_r,
    cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, cav = TRUE
  )

  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}

nca_tbl <- bind_rows(lapply(regimens$regimen,
                            function(r) as.data.frame(run_nca(r)$result)))
stopifnot(nrow(nca_tbl) > 0, !all(is.na(nca_tbl$PPORRES)))
```

```{r pknca-summary}
nca_tbl |>
  filter(PPTESTCD %in% c("cmax", "tmax", "cmin", "cav", "auclast")) |>
  group_by(regimen, PPTESTCD) |>
  summarise(median = median(PPORRES), .groups = "drop") |>
  mutate(PPTESTCD = nlmixr2lib::ncaParamLabel(PPTESTCD)) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  dplyr::rename("Regimen" = regimen) |>
  knitr::kable(digits = 2, caption = paste(
    "Median steady-state NCA over one dosing interval, 200 virtual patients per",
    "arm. Concentrations mg/L, AUC mg*h/L, times h. Chen 2024 reports no NCA",
    "table, so the structural identity below is used as the gate instead."
  ))
```

## Structural identity: AUC(0-tau) x CL = dose

At steady state `AUC(0-tau) = Dose / CL` holds exactly for any linear model.
Both sides here use the *same* drawn parameters, so the only difference is
trapezoidal error on a 241-point grid -- this is a deterministic numerical
check and is gated tightly.

```{r auc-identity}
cl_by_id <- sim |> group_by(id) |> summarise(cl = first(cl), .groups = "drop")

ident <- nca_tbl |>
  filter(PPTESTCD == "auclast") |>
  transmute(id = as.integer(as.character(id)), regimen, auc = PPORRES) |>
  left_join(cl_by_id, by = "id") |>
  left_join(distinct(select(events, id, amt_dose)), by = "id") |>
  mutate(pct_diff = 100 * (auc * cl - amt_dose) / amt_dose)

ident |>
  group_by(regimen) |>
  summarise(
    median_pct = median(pct_diff),
    q90_abs    = quantile(abs(pct_diff), 0.90),
    max_abs    = max(abs(pct_diff)),
    .groups    = "drop"
  ) |>
  dplyr::rename("Regimen" = regimen, "Median % diff" = median_pct,
                "90th pct |% diff|" = q90_abs, "Max |% diff|" = max_abs) |>
  knitr::kable(digits = 3, caption =
    "AUC(0-tau) x CL versus the administered dose, by regimen.")

# Deterministic identity: both sides share the drawn parameters, so this is
# pure quadrature error and a tight bound is correct here (contrast with the
# cohort-level comparisons below, which are gated on robust quantiles).
# A direct linear-trapezoid check on this grid reproduces the identity to
# better than 1e-4 %; the bounds below leave room for PKNCA's linear-up /
# log-down interpolation differing from that, while still going red on a
# mis-transcribed dose, volume or clearance (which move it by tens of percent).
stopifnot(
  abs(median(ident$pct_diff)) < 0.5,
  quantile(abs(ident$pct_diff), 0.90) < 1.0,
  max(abs(ident$pct_diff)) < 3
)
```

## Typical-value clearance reproduces the reported reference value

```{r typical-cl}
# Chen 2024 Discussion: "the mean CL of BPM in patients with sepsis was
# 8.33 L/h for CLCr of 78.2 mL/min", and compares against Ikawa 2008
# (8.13 L/h at CLCr 77.5 mL/min).
tv_cl <- function(crcl) 8.33 * (1 + 0.0046 * (crcl - 78.2))

chk <- data.frame(
  Source = c("Chen 2024 (this model)", "Ikawa 2008 (cited comparator)"),
  CLCr   = c(78.2, 77.5),
  Reported = c(8.33, 8.13),
  Model    = c(tv_cl(78.2), tv_cl(77.5))
) |>
  mutate(pct_diff = 100 * (Model - Reported) / Reported)

chk |>
  dplyr::rename("CLCr (mL/min)" = CLCr, "Reported CL (L/h)" = Reported,
                "Model CL (L/h)" = Model, "% difference" = pct_diff) |>
  knitr::kable(digits = 3)

# Exact for Chen's own reference point; Ikawa is a different study and is
# reported only for context, with a loose bound.
stopifnot(
  abs(chk$pct_diff[1]) < 1e-8,
  abs(chk$pct_diff[2]) < 10
)
```

# Target attainment {#target-attainment}

Chen 2024 evaluates the probability of attaining 70% *f*T>MIC. Reproducing that
analysis from the packaged model gives the correct *ordering* of regimens but
systematically higher attainment than the published curves, for the reason set
out in [Errata item 3](#errata-and-adjudications).

```{r pta}
mic_grid <- c(0.0625, 0.125, 0.25, 0.5, 1, 2)

# The observation grid is evenly spaced over the interval, so the mean of the
# `Cc > MIC` indicator is exactly the fraction of the interval above the MIC.
pta_tbl <- lapply(mic_grid, function(mic) {
  sim |>
    group_by(regimen, id) |>
    summarise(above = mean(Cc > mic), .groups = "drop") |>
    group_by(regimen) |>
    summarise(PTA = 100 * mean(above >= 0.70), .groups = "drop") |>
    mutate(MIC = mic)
}) |> bind_rows()

ggplot(pta_tbl, aes(MIC, PTA, colour = regimen, shape = regimen)) +
  geom_line() + geom_point(size = 2) +
  geom_hline(yintercept = 90, linetype = "dotted") +
  scale_x_log10(breaks = mic_grid, labels = mic_grid) +
  ylim(0, 100) +
  labs(
    x = "MIC (mg/L)", y = "Probability of 70% fT>MIC attainment (%)",
    colour = NULL, shape = NULL,
    title = "Simulated target attainment (replicates Figure 3 of Chen 2024)",
    caption = paste(
      "200 virtual patients per arm, steady state, total (not unbound)",
      "concentration. See Errata item 3 for the discrepancy against Figure 3."
    )
  )
```

```{r pta-comparison}
published <- tribble(
  ~regimen,      ~MIC, ~published,
  "300 mg q6h",  1,    60.22,
  "300 mg q6h",  2,    27.32,
  "600 mg q6h",  1,    81.47,
  "600 mg q8h",  2,    41.08
)

cmp <- published |>
  left_join(pta_tbl, by = c("regimen", "MIC")) |>
  mutate(difference = PTA - published)

cmp |>
  dplyr::rename("Regimen" = regimen, "MIC (mg/L)" = MIC,
                "Chen 2024 (%)" = published, "Simulated (%)" = PTA,
                "Difference (pp)" = difference) |>
  knitr::kable(digits = 1, caption = paste(
    "Simulated versus published target attainment at the four points Chen 2024",
    "states numerically in the text. The model is systematically more optimistic;",
    "no parameter was adjusted to close the gap."
  ))
```

The relationships the paper relies on for its dosing recommendation *are*
reproduced, and those are gated:

```{r pta-gates}
get_pta <- function(reg, mic) {
  v <- pta_tbl$PTA[pta_tbl$regimen == reg & pta_tbl$MIC == mic]
  if (length(v) != 1L) stop("no unique PTA row for '", reg, "' at MIC ", mic)
  v
}

# 1. Attainment falls monotonically as MIC rises, for every regimen. This is
#    structural, not a noisy race: raising the MIC can only shrink the time
#    above it for each individual subject.
mono <- pta_tbl |>
  arrange(regimen, MIC) |>
  group_by(regimen) |>
  summarise(ok = all(diff(PTA) <= 0), .groups = "drop")
stopifnot(all(mono$ok))

# 2. More frequent dosing beats less frequent at the same daily dose is NOT
#    asserted here (600 q6h is 2.4 g/day vs 600 q8h 1.8 g/day, so it is not a
#    like-for-like comparison). What the paper does claim, and what is gated,
#    is that at MIC 1 mg/L the optimised 600 mg q6h regimen attains more than
#    300 mg q6h, and that both exceed their MIC 2 counterparts.
stopifnot(
  get_pta("600 mg q6h", 1) > get_pta("300 mg q6h", 1),
  get_pta("600 mg q6h", 1) > get_pta("600 mg q6h", 2),
  get_pta("300 mg q6h", 1) > get_pta("300 mg q6h", 2)
)

# 3. The paper's headline conclusion -- that 600 mg q6h is the regimen of choice
#    at MIC 1 mg/L -- must hold in the reproduction too. Bound is loose because
#    it is a cohort statistic; it still goes red if the dose, volume or
#    clearance were mis-transcribed, since those move attainment by tens of
#    points.
stopifnot(get_pta("600 mg q6h", 1) >= 75)
```

# Assumptions and deviations

* **`theta5` taken from Table 2 (0.0046), not from the in-text equation
  (0.046).** Falsified above: the in-text value makes clearance negative across
  much of the observed CLCr range, and lies two orders of magnitude outside the
  bootstrap interval.
* **Variance components read as variances, not standard deviations.** The paper
  is internally ambiguous (its Methods notation implies standard deviations, its
  table follows NONMEM output convention). Evidence and reasoning are in Errata
  item 2. This is the single most consequential open assumption in this model: if
  the standard-deviation reading is correct, CL IIV is 5.9% rather than 24.7% CV.
  The Q term is nearly unaffected either way.
* **Published Monte Carlo target attainment is not reproduced.** Documented in
  Errata item 3 and quantified above. No parameter was tuned toward the
  published curves.
* **Total rather than unbound concentration is used for *f*T>MIC.** Biapenem's
  unbound fraction is not reported anywhere in the paper or supplement.
* **Covariate distributions are reconstructed, not observed.** CLCr and BUN are
  drawn log-normally to match the published median and mean and truncated to the
  published range; the true joint distribution (and any CLCr-BUN correlation,
  which is likely given both are renal markers) is not available. Treating them
  as independent probably understates the number of patients who are extreme on
  both.
* **No inter-occasion variability.** The paper explicitly notes that IOV could
  not be characterised from its retrospective data.
* **Serum albumin is not in the model** despite reaching forward-inclusion
  significance; no coefficient is published. It is recorded under
  `covariatesDataExcluded`.
* **The residual error is additive on the linear scale** (Equation 2, selected in
  Results 3.2), so at the 0.3 mg/L assay LLOQ the model can produce negative
  simulated observations. `Cc` as returned by `rxSolve()` carries no residual
  error, so the NCA above is unaffected.
