---
title: "Overall-survival MBMA in platinum-ineligible NSCLC (Chen 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Overall-survival MBMA in platinum-ineligible NSCLC (Chen 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(ggplot2)
```

```{r model-load, include = FALSE}
mod_fn <- readModelDb("Chen_2026_nsclc_os_mbma")
mod_meta <- mod_fn()
mod <- rxode2::rxode(mod_fn)
```

* Citation: `r mod_meta$reference`
* Article: [CPT Pharmacometrics Syst Pharmacol. 2026;15:e70197](https://doi.org/10.1002/psp4.70197)

# Model and source

Chen 2026 asks a question that model-based meta-analysis (MBMA) is rarely used
for: not whether the *investigational* arm of a registrational trial works, but
whether its *control* arm was a fair comparator. The phase III IPSOS trial
(NCT03191786) compared atezolizumab monotherapy against single-agent
chemotherapy in treatment-naive advanced non-small cell lung cancer (NSCLC)
patients unsuitable for platinum-doublet chemotherapy, but its protocol
restricted the control arm to gemcitabine or vinorelbine. Both the NCCN and the
ESMO guidelines list a broader set of single agents for this population, so the
authors assembled a literature database of historical single-agent trials and
asked whether the IPSOS control arm was non-inferior to that broader set.

The model is a mixed-effects **non-parametric conditional-probability** survival
model, fitted in NONMEM 7.5.1 by the Laplacian method to digitised
Kaplan-Meier curves. Its two moving parts are:

1. A **reference survival curve** with no assumed shape. The logit of the
   reference survival probability is an estimated monotone random walk over 30
   knots (months 1-24, then 27, 30, 33, 36, 42 and 48). The reference is a
   historical single-agent-chemotherapy control arm composed entirely of ECOG
   performance status 0 patients.
2. A **proportional rescaling** of that curve per study arm,
   `S_arm(t) = S_ref(t)^exp(log(HR))`, where `log(HR)` is a linear combination
   of covariates plus a between-trial random effect. Because `S = exp(-H)`,
   raising the survival function to the power `exp(log(HR))` multiplies the
   cumulative hazard by the hazard ratio, so this is an ordinary
   proportional-hazards model written on the survival scale.

This file encodes the paper's **final model, Model 010**, whose complete
parameter set is Table S8 of the Supporting Information.

```{r model-print}
mod
```

# Population

```{r population, echo = FALSE}
knitr::kable(
  data.frame(
    Field = c("Species", "Studies", "Arms", "Participants", "Disease",
              "Age", "Sex", "Region", "Treatments"),
    Value = c(
      mod_meta$population$species,
      format(mod_meta$population$n_studies),
      format(mod_meta$population$n_arms),
      format(mod_meta$population$n_subjects),
      "Treatment-naive locally advanced or metastatic NSCLC, unsuitable for platinum-doublet chemotherapy",
      "Arm-level median 74.0 years (range across arms 61.0 to 79.0)",
      "Arm-level median 74.7% male (range 38.9 to 94.2)",
      "Multinational; arm-level Asian enrolment median 0%, range 0 to 100",
      "Single-agent docetaxel, gemcitabine, paclitaxel, pemetrexed or vinorelbine"
    ),
    stringsAsFactors = FALSE
  ),
  caption = "Analysis population (Chen 2026 Tables 1 and 2)."
)
```

Eligibility for the database required ECOG performance status `>= 2`, **or**
age `>= 70` years, **or** platinum ineligibility as defined by the respective
published study. Arm-level performance-status composition is the covariate that
carries the whole analysis: median 14.7% ECOG 0, 42.9% ECOG 1 and 28.8% ECOG 2
or 3 across the 41 arms.

Model 010 is fitted with the IPSOS **atezolizumab** arm excluded (Chen 2026
Table 5), leaving 3335 participants across 40 arms.

# Source trace

Every value in `ini()` comes from Table S8 ("Final Model Parameters (mod010)")
of the Supporting Information, except the final reference-curve decrement,
which was fixed to zero and therefore does not appear there.

```{r source-trace, echo = FALSE}
knitr::kable(
  data.frame(
    Quantity = c(
      "logit S_ref(1 month)",
      "29 logit decrements, months 2-48",
      "Decrement for months 42 to 48",
      "ECOG PS 1 effect on log(HR)",
      "ECOG PS 2 or 3 effect on log(HR)",
      "IPSOS control arm effect on log(HR)",
      "Between-trial SD on log(HR)",
      "Random-walk / proportional-hazards structure",
      "Knot grid (months 1-24, 27, 30, 33, 36, 42, 48)",
      "Arm-level ECOG percentages",
      "IPSOS control arm identity"
    ),
    `Source location` = c(
      "Table S8 row 1 ('logit(Survival at 1-mo)') = 2.802",
      "Table S8 rows 2-29",
      "Control stream psp470197-sup-0002-Supinfo2.ctl, $THETA row 30: '0 FIX ; TIME48'",
      "Table S8 'ECOG1 effect' = 0.004; Table 3 Model 010 ECOG1HR = 1.004",
      "Table S8 'ECOG2,3 effect' = 0.769; Table 3 Model 010 ECOG2,3HR = 2.158",
      "Table S8 'IPSOS control arm effect' = -0.633; Table 3 Model 010 IPSOSHR = 0.531",
      "Table S8 'Between-trial SD' = 0.257; $OMEGA is '1 FIX' so this THETA is an SD",
      "Control stream $PRED block; Methods 2.4 and Figure 1",
      "TIME column of the distributed dataset psp470197-sup-0001-Supinfo1.csv",
      "Tables S5 (raw) and S6 (with imputations); dataset columns PS0-PS3",
      "Dataset ID 15 ARM 2 (Lee 2023, gemcitabine or vinorelbine, n = 151)"
    ),
    check.names = FALSE,
    stringsAsFactors = FALSE
  ),
  caption = "Source trace for the model structure and every ini() value."
)
```

Table S8 is internally consistent: its printed RSE column reproduces as
`SE / Value * 100` for every row, which is how the transcription above was
checked.

# Reference survival curve

The first validation is closed-form and deterministic. The model's `surv0`
output at each knot must equal the inverse logit of the Table S8 cumulative
sum, with no simulation error and no random draw involved.

```{r ref-curve}
knots <- c(1:24, 27, 30, 33, 36, 42, 48)

# Table S8 rows 1-29, transcribed independently of the model file, plus the
# month 42-48 decrement fixed to zero in the control stream.
lgtsurv_01 <- 2.802
decrements <- c(
  0.831, 0.520, 0.401, 0.301, 0.258, 0.189, 0.196, 0.159, 0.152, 0.143,
  0.114, 0.099, 0.168, 0.102, 0.098, 0.094, 0.083, 0.115, 0.187, 0.051,
  0.093, 0.105, 0.135, 0.137, 0.225, 0.260, 0.442, 0.517, 0.000
)
stopifnot(length(decrements) == length(knots) - 1L)

expected_lgt  <- lgtsurv_01 - cumsum(c(0, decrements))
expected_surv <- plogis(expected_lgt)

# Solve the packaged model on an all-ECOG-0 historical control arm, which is
# the reference by construction, at the typical value of the between-trial
# random effect.
ref_arm <- data.frame(
  id = 1L, time = knots, evid = 0L, amt = NA_real_,
  PS_ECOG_1_PCT = 0, PS_ECOG_2_PCT = 0, PS_ECOG_3_PCT = 0,
  TRT_IPSOS_CONTROL = 0
)
ref_sim <- rxode2::rxSolve(
  rxode2::zeroRe(mod), ref_arm, returnType = "data.frame"
)

stopifnot(nrow(ref_sim) == length(knots))
max_abs_err <- max(abs(ref_sim$surv0 - expected_surv))
max_abs_err

# Deterministic identity: the two sides are the same arithmetic, so this is
# pure floating-point error and a tight bound is correct here. It goes red on
# any mis-transcribed decrement, a knot placed at the wrong month, or a sign
# error in the random walk.
stopifnot(max_abs_err < 1e-10)

# The reference arm must have hazard ratio exactly 1 -- that is what makes it
# the reference.
stopifnot(max(abs(ref_sim$hr - 1)) < 1e-12)
stopifnot(max(abs(ref_sim$surv - ref_sim$surv0)) < 1e-12)
```

```{r ref-curve-table, echo = FALSE}
knitr::kable(
  data.frame(
    Month = knots,
    `logit S_ref` = round(expected_lgt, 4),
    `S_ref (Table S8)` = round(expected_surv, 4),
    `S_ref (model)` = round(ref_sim$surv0, 4),
    check.names = FALSE
  ),
  caption = "Reference survival curve: Table S8 arithmetic against the packaged model."
)
```

The reference curve crosses 50% survival between months 8 and 9, giving a
median overall survival of roughly 8.7 months for a historical
single-agent-chemotherapy arm made up entirely of ECOG 0 patients.

```{r ref-median}
# Interpolate over the strictly decreasing part of the curve only. The final
# decrement is fixed to zero, so S_ref(42) and S_ref(48) are tied and would
# make the inverse interpolation ambiguous.
strict <- knots <= 42
median_os_ref <- approx(x = expected_surv[strict], y = knots[strict], xout = 0.5)$y
round(median_os_ref, 2)
stopifnot(!is.na(median_os_ref), median_os_ref > 8, median_os_ref < 9)
```

# Hazard ratios: reproducing Table 3

The three covariate effects exponentiate to the hazard ratios Chen 2026 prints
for Model 010 in Table 3. The comparison below is against values typed from the
paper, not against the model's own inputs, so it can go red.

```{r hazard-ratios}
ini_df <- as.data.frame(mod$iniDf)
est <- function(nm) ini_df$est[match(nm, ini_df$name)]

hr_tab <- data.frame(
  Effect = c("ECOG PS 1 vs ECOG PS 0",
             "ECOG PS 2 or 3 vs ECOG PS 0",
             "IPSOS control vs historical control"),
  `Model HR` = round(exp(c(est("e_ecog1"), est("e_ecog23"), est("e_ipsosctl"))), 3),
  `Published HR (Table 3, Model 010)` = c(1.004, 2.158, 0.531),
  `Published 95% CI` = c("0.412 to 2.443", "1.003 to 4.639", "0.424 to 0.664"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
knitr::kable(hr_tab, caption = "Model-implied hazard ratios against Chen 2026 Table 3.")

stopifnot(all(abs(hr_tab$`Model HR` - hr_tab$`Published HR (Table 3, Model 010)`) < 5e-4))
```

The between-trial standard deviation is stored as a variance, so it must be
recovered by taking a square root before comparing with the Table S8 value
of 0.257.

```{r between-trial-sd}
omega_var <- mod$omega[["eta_study", "eta_study"]]
between_trial_sd <- sqrt(omega_var)
round(between_trial_sd, 3)
stopifnot(abs(between_trial_sd - 0.257) < 5e-4)
```

# Virtual cohort: the study arms

The "subjects" of this model are study arms, each carrying its arm-level ECOG
composition. Two arms are of particular interest because they used the *same*
chemotherapy regimen (gemcitabine or vinorelbine, investigator's choice): the
IPSOS control arm and O'Brien 2008. Chen 2026 remarks on the gap between them
in Section 3.2.3.

Arm-level covariates are taken from the distributed analysis dataset
(`psp470197-sup-0001-Supinfo1.csv`), which matches Chen 2026 Table S6.

```{r arms}
arms <- data.frame(
  id    = 1:3,
  label = c("Reference (historical control, all ECOG 0)",
            "IPSOS control arm (Lee 2023, n = 151)",
            "O'Brien 2008 (n = 190)"),
  PS_ECOG_1_PCT     = c(0,  11.3,   0),
  PS_ECOG_2_PCT     = c(0,  76.8, 100),
  PS_ECOG_3_PCT     = c(0,  10.6,   0),
  TRT_IPSOS_CONTROL = c(0,   1,     0),
  stringsAsFactors  = FALSE
)
knitr::kable(arms[, -1], caption = "Arm-level covariates entering log(HR).")
```

Chen 2026 explains the difference between the unadjusted and adjusted hazard
ratios by noting that "87% of IPSOS patients had ECOG PS 2 or 3". That figure
is reproducible from the arm composition.

```{r ipsos-ecog}
ipsos_ecog23 <- with(arms[arms$id == 2, ], PS_ECOG_2_PCT + PS_ECOG_3_PCT)
ipsos_ecog23
# Paper states 87%; the dataset gives 87.4%.
stopifnot(abs(ipsos_ecog23 - 87) < 1)
```

Against a database median of 28.8% ECOG 2 or 3 (Table 2), the IPSOS control arm
is an extreme in performance-status burden. That is precisely why the paper's
unadjusted estimate (Model 007a, HR 0.812) and its ECOG-adjusted estimate
(Model 010, HR 0.531) differ so much: without the adjustment, the IPSOS control
arm's sicker population is charged against the treatment.

The size of that correction is recoverable from the model. What moves the
estimate is not the IPSOS arm's ECOG burden in absolute terms, but its burden
*relative to the typical historical arm* -- an unadjusted model charges the
IPSOS control arm only for the difference, because the historical arms carry a
performance-status burden of their own.

```{r ecog-adjustment}
ecog_loghr <- function(pct_ps1, pct_ps23) {
  est("e_ecog1") * pct_ps1 / 100 + est("e_ecog23") * pct_ps23 / 100
}

# The IPSOS control arm, from its own composition.
loghr_ipsos <- ecog_loghr(
  arms$PS_ECOG_1_PCT[2],
  arms$PS_ECOG_2_PCT[2] + arms$PS_ECOG_3_PCT[2]
)

# The typical historical arm, from the database medians in Chen 2026 Table 2:
# 42.9% ECOG PS 1 and 28.8% ECOG PS 2 or 3.
loghr_typical <- ecog_loghr(42.9, 28.8)

relative_ecog_hr <- exp(loghr_ipsos - loghr_typical)
round(relative_ecog_hr, 3)

# Model 007a made no ECOG adjustment and estimated HR 0.812; Model 010 adjusts
# and estimates 0.531. The ratio of the two published estimates is the size of
# the correction the ECOG composition had to supply.
published_ratio <- 0.812 / 0.531
round(published_ratio, 3)

# Both sides are fixed numbers -- no simulation, no random draw -- but the two
# models also differ in their fitted reference curves, so exact agreement is
# not expected. The bound admits that slack while still going red on a wrong
# ECOG coefficient (halving e_ecog23 moves the left-hand side to 1.24).
stopifnot(abs(relative_ecog_hr - published_ratio) < 0.15)
```

# Predicted versus observed survival curves

The digitised Kaplan-Meier data distributed with the paper let the packaged
model be checked against the curves it was fitted to.

**Time alignment.** In the distributed dataset each record spans the interval
`(TLAST, TIME]`: `SURV` is the survival probability at the *start* of the
interval and `NDIE` the deaths within it. This is verifiable from the data
itself, since `NALIVE` equals `N * SURV` row by row. The observed survival
probability therefore pairs with `TLAST`, and that is the alignment used below.

```{r observed}
observed <- dplyr::bind_rows(
  data.frame(
    label = "IPSOS control arm (Lee 2023, n = 151)",
    time  = c(0:24, 27),
    surv  = c(1.000, 0.963, 0.925, 0.841, 0.739, 0.666, 0.577, 0.565, 0.526,
              0.512, 0.474, 0.430, 0.381, 0.359, 0.320, 0.297, 0.273, 0.256,
              0.240, 0.226, 0.204, 0.187, 0.173, 0.134, 0.127, 0.104),
    stringsAsFactors = FALSE
  ),
  data.frame(
    label = "O'Brien 2008 (n = 190)",
    time  = 0:18,
    surv  = c(1.000, 0.939, 0.908, 0.843, 0.773, 0.657, 0.539, 0.456, 0.406,
              0.342, 0.308, 0.276, 0.259, 0.235, 0.217, 0.196, 0.171, 0.146,
              0.136),
    stringsAsFactors = FALSE
  )
)
```

```{r predicted}
grid <- seq(0, 48, by = 0.25)
ev <- do.call(rbind, lapply(seq_len(nrow(arms)), function(i) {
  data.frame(
    id = arms$id[i], time = grid, evid = 0L, amt = NA_real_,
    PS_ECOG_1_PCT     = arms$PS_ECOG_1_PCT[i],
    PS_ECOG_2_PCT     = arms$PS_ECOG_2_PCT[i],
    PS_ECOG_3_PCT     = arms$PS_ECOG_3_PCT[i],
    TRT_IPSOS_CONTROL = arms$TRT_IPSOS_CONTROL[i]
  )
}))

sim <- rxode2::rxSolve(rxode2::zeroRe(mod), ev, returnType = "data.frame")
if (is.null(sim$id)) sim$id <- 1L
sim$id <- as.integer(as.character(sim$id))
sim <- dplyr::left_join(sim, arms[, c("id", "label")], by = "id")
stopifnot(!anyNA(sim$label), nrow(sim) == nrow(arms) * length(grid))

arm_hr <- sim |>
  dplyr::distinct(label, hr) |>
  dplyr::mutate(hr = round(hr, 3))
knitr::kable(arm_hr, caption = "Model-implied hazard ratio per arm, relative to the reference curve.")
```

```{r fig-curves, fig.cap = "Model-predicted overall-survival curves (lines) against the digitised Kaplan-Meier data the model was fitted to (points). Replicates the per-trial fits of Chen 2026 Figure S2."}
ggplot2::ggplot(sim, ggplot2::aes(time, surv, colour = label)) +
  ggplot2::geom_step(linewidth = 0.7) +
  ggplot2::geom_point(
    data = observed, ggplot2::aes(time, surv, colour = label),
    inherit.aes = FALSE, size = 1.6
  ) +
  ggplot2::coord_cartesian(xlim = c(0, 30), ylim = c(0, 1)) +
  ggplot2::labs(
    x = "Time (months)", y = "Overall survival probability", colour = NULL
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom", legend.direction = "vertical")
```

The IPSOS control arm -- the arm the whole analysis is about -- is described
closely by the typical-value prediction.

```{r gate-ipsos}
pred_at <- function(lbl, times) {
  s <- sim[sim$label == lbl, ]
  idx <- match(times, s$time)
  stopifnot(!anyNA(idx))
  s$surv[idx]
}

ipsos_obs <- observed[observed$label == "IPSOS control arm (Lee 2023, n = 151)", ]
ipsos_chk <- dplyr::mutate(
  ipsos_obs,
  pred = pred_at("IPSOS control arm (Lee 2023, n = 151)", time),
  diff = pred - surv
)
stopifnot(nrow(ipsos_chk) == 26L)   # the gate must have rows to test

summary_stats <- c(
  median_abs_diff = median(abs(ipsos_chk$diff)),
  q90_abs_diff    = unname(quantile(abs(ipsos_chk$diff), 0.9)),
  max_abs_diff    = max(abs(ipsos_chk$diff))
)
round(summary_stats, 4)

# Deterministic comparison (typical-value solve against fixed published data --
# no random draw on either side), so these bounds are reproducible across
# machines. They go red on a mis-transcribed decrement, a wrong ECOG
# coefficient, or a lost IPSOS indicator, each of which moves the curve by
# well over 0.05 in survival probability.
stopifnot(
  median(abs(ipsos_chk$diff)) < 0.03,
  quantile(abs(ipsos_chk$diff), 0.9) < 0.05
)
```

O'Brien 2008 is a different story, and the paper says so. Chen 2026
Section 3.2.3 notes that "the comparison between O'Brien and the IPSOS control
shows that the IPSOS control was superior to O'Brien, despite having the same
chemotherapy regimen of gemcitabine or vinorelbine", and raises publication era
or unmeasured heterogeneity as candidate explanations. In this model that gap
is absorbed by the between-trial random effect: the arm survives markedly
longer than its imputed all-ECOG-2 composition predicts.

```{r gate-obrien}
obrien_obs <- observed[observed$label == "O'Brien 2008 (n = 190)", ]
obrien_chk <- dplyr::mutate(
  obrien_obs,
  pred = pred_at("O'Brien 2008 (n = 190)", time),
  diff = pred - surv
)
stopifnot(nrow(obrien_chk) == 19L)

# The typical-value prediction sits BELOW the observed curve throughout -- the
# model over-predicts this arm's hazard.
stopifnot(all(obrien_chk$diff[obrien_chk$time > 0] < 0))

# Size of the between-trial random effect required to reconcile the arm,
# estimated over months 6 to 18 where the curves are well separated.
w <- obrien_chk$time >= 6 & obrien_chk$time <= 18
eta_needed <- median(
  log(log(obrien_chk$surv[w]) / log(pred_at("Reference (historical control, all ECOG 0)",
                                            obrien_chk$time[w]))) - est("e_ecog23")
)
round(eta_needed, 3)
round(eta_needed / between_trial_sd, 2)

# Roughly a 2-standard-deviation trial: unusual but inside the between-trial
# distribution the model estimates, and consistent with the paper singling this
# study out. A loose bound, because the claim being made is only "large but not
# impossible".
stopifnot(eta_needed < 0, abs(eta_needed / between_trial_sd) > 1,
          abs(eta_needed / between_trial_sd) < 3.5)
```

# Between-trial prediction interval

Chen 2026 Figure 3 compares the IPSOS control arm against the *prediction
range* of historical trials. The equivalent here is the between-trial
distribution of the reference curve: `log(HR)` is normal with standard
deviation 0.257, so the 90% prediction band for a new historical
all-ECOG-0 trial arm is available in closed form and needs no simulation.

```{r band}
z90 <- qnorm(0.95)
band <- data.frame(
  time = grid,
  ref  = pred_at("Reference (historical control, all ECOG 0)", grid)
) |>
  dplyr::mutate(
    lower = ref^exp( z90 * between_trial_sd),
    upper = ref^exp(-z90 * between_trial_sd)
  )

ipsos_pred <- data.frame(
  time = grid,
  surv = pred_at("IPSOS control arm (Lee 2023, n = 151)", grid)
)

# Ordering is structural, not stochastic: exp(+z*sd) > 1 shortens survival.
stopifnot(all(band$lower <= band$ref + 1e-12),
          all(band$upper >= band$ref - 1e-12))
```

```{r fig-band, fig.cap = "IPSOS control arm against the 90% between-trial prediction band for a historical all-ECOG-0 control arm. Analogous to Chen 2026 Figure 3."}
ggplot2::ggplot(band, ggplot2::aes(time)) +
  ggplot2::geom_ribbon(
    ggplot2::aes(ymin = lower, ymax = upper),
    fill = "grey70", alpha = 0.5
  ) +
  ggplot2::geom_step(ggplot2::aes(y = ref, linetype = "Historical reference (ECOG 0)"),
                     linewidth = 0.7) +
  ggplot2::geom_step(
    data = ipsos_pred,
    ggplot2::aes(time, surv, linetype = "IPSOS control arm"),
    linewidth = 0.7, colour = "firebrick"
  ) +
  ggplot2::coord_cartesian(xlim = c(0, 36), ylim = c(0, 1)) +
  ggplot2::labs(x = "Time (months)", y = "Overall survival probability",
                linetype = NULL) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

A simulated cohort of trial arms reproduces the same band, and confirms that
`eta_study` is wired into `log(HR)` rather than being inert.

```{r sim-cohort}
n_arms <- 200L
rxode2::rxSetSeed(20260912)
cohort_ev <- do.call(rbind, lapply(seq_len(n_arms), function(i) {
  data.frame(
    id = i, time = knots, evid = 0L, amt = NA_real_,
    PS_ECOG_1_PCT = 0, PS_ECOG_2_PCT = 0, PS_ECOG_3_PCT = 0,
    TRT_IPSOS_CONTROL = 0
  )
}))
cohort <- rxode2::rxSolve(mod, cohort_ev, returnType = "data.frame")

# The random effect must actually vary between arms.
arm_log_hr <- cohort |>
  dplyr::distinct(id, lhr) |>
  dplyr::pull(lhr)
stopifnot(length(arm_log_hr) == n_arms, stats::sd(arm_log_hr) > 0)

round(c(mean = mean(arm_log_hr), sd = stats::sd(arm_log_hr)), 3)

# A 200-draw sample standard deviation has about a 5% relative standard error,
# so this is a deliberately wide envelope around the nominal 0.257 rather than
# a bound tightened to one observed run (see the CI-reproducibility note in
# the package vignette conventions).
stopifnot(abs(mean(arm_log_hr)) < 0.1,
          stats::sd(arm_log_hr) > 0.18, stats::sd(arm_log_hr) < 0.36)
```

# Assumptions and deviations

* **Between knots the curve is held flat.** The source likelihood evaluates the
  reference curve only at the 30 knot times, so the model as fitted says
  nothing about intermediate times. The packaged model holds `surv0` constant
  from each knot until the next, which is both the conservative reading and
  what a Kaplan-Meier curve is. Times between 24 and 27 months therefore report
  the 24-month value, and so on for the wider late knots.
* **Support is 1 to 48 months.** The month 42-48 decrement was fixed to zero in
  the source control stream, so the reference curve is flat over that span and
  the model carries no information beyond 48 months. Extrapolating past 48
  months returns the 42-month value and should not be treated as a prediction.
* **The random effect is between-trial, not between-subject.** `eta_study` is
  an MBMA variance component describing how far one trial arm's whole survival
  curve sits from the model. Simulating it produces a distribution of *trial
  arms*, not of patients; this model cannot generate individual-subject event
  times.
* **No residual error is encoded.** Chen 2026 maximises a user-supplied
  binomial `-2LL` over per-interval death counts, with no `$SIGMA` in the
  control stream, so there is no residual-error parameter to carry and no
  observation endpoint is declared. The weight column `WT` used in that
  likelihood is a fitting device and has no role in simulation.
* **ECOG 2 and ECOG 3 share one coefficient**, matching
  `COVECOG2*(PS2+PS3)/100` in the source control stream. The two are kept as
  separate covariate columns because the source dataset carries them apart.
* **Arm-level ECOG percentages are partly imputed.** Chen 2026 imputed 30 ECOG
  percentages, first by mapping Karnofsky Performance Scores and otherwise by
  splitting composite ECOG categories with a logistic regression calibrated on
  studies with similar entry criteria. O'Brien 2008 carries an imputed 100%
  ECOG 2, which is the largest single driver of its predicted hazard ratio and
  a plausible contributor to the misfit shown above.
* **Only Model 010 is packaged.** See Errata for why the paper's other models
  are not reproducible from the published material.
* **Digitisation and censoring.** The database was built by digitising
  published Kaplan-Meier curves at monthly intervals with Engauge Digitizer
  12.1. Censoring tick marks could not be recovered, which the authors note may
  over-weight later time points; they mitigated it by trimming curve tails
  where fewer than 10 patients remained at risk or survival fell below 0.1.

# Errata

* **The distributed control stream is not Model 010.** The Supporting
  Information file `psp470197-sup-0002-Supinfo2.ctl` is referenced by the
  Figure S2 caption as "the control file ... for the final model (model010)",
  but its `$PRED` block fixes the ECOG effects to the external Lee
  meta-analysis values (`LOG(1.43)` and `LOG(2.71)`), adds fixed Stage IV and
  sex effects that appear in no reported model, and carries `ATEZO` as
  `0 FIX` with no IPSOS control term. Model 010 by definition *estimates* the
  ECOG effects (Table 3) and retains an IPSOS control effect. The file is
  therefore a different run -- structurally closest to Model 007b. It is used
  here only as the authority for the **model form** (the logit random walk, the
  `S^exp(log HR)` rescaling, the conditional-death-probability likelihood, and
  the `0 FIX` final decrement); every **parameter value** comes from Table S8,
  which is explicitly headed "Final Model Parameters (mod010)" and which
  reproduces all three of the Table 3 Model 010 hazard ratios exactly.
* **The paper's headline hazard ratios come from Model 041, not Model 010.**
  The Abstract, Discussion and Conclusions quote HR 0.543 (95% CI 0.435 to
  0.677) for the IPSOS control arm and HR 0.418 (95% CI 0.335 to 0.522) for the
  IPSOS atezolizumab arm. Both are Model 041 (Table 4), which adds the
  atezolizumab arm to the dataset and estimates a second treatment effect.
  Model 010's corresponding IPSOS control estimate is 0.531 (95% CI 0.424 to
  0.664). Only Model 010 has a published reference survival curve -- Table S8 is
  the paper's only complete parameter listing -- so **the atezolizumab effect is
  not packaged**. Grafting the Model 041 atezolizumab hazard ratio onto the
  Model 010 reference curve would mix parameters from two different fits, and is
  deliberately not done.
* **Abstract-to-Discussion inconsistency in the IPSOS control confidence
  interval.** The Abstract reports 0.543 (95% CI 0.435 to 0.677) while Table 4
  reports the same point estimate with 95% CI 0.436 to 0.676. The discrepancy is
  in the last digit of each bound and does not affect the model.
* **Table S8 abbreviation gloss.** The Table S8 footnote expands RSE as
  "residual standard error", but its own Note defines the column as
  `SE / Estimate * 100%`, i.e. the *relative* standard error. The values in the
  column are relative standard errors; the abbreviation list is wrong.
* **Table 4 is unreadable in the PMC text conversion.** The open-access PMC
  rendering of Table 4 collapses its columns into a single cell, losing the
  model-to-hazard-ratio mapping. The values used here were read from the
  laid-out PDF, where the table is intact.
* **Supporting Information acquisition.** The Supporting Information was not
  supplied with the task and was retrieved from the Europe PMC supplementary
  files endpoint for PMC12909276. It comprises `Supinfo1` (the NONMEM analysis
  dataset), `Supinfo2` (the control stream discussed above) and `Supinfo3`
  (Tables S1 to S8 and Figures S1 to S3).
