---
title: "Isoniazid + acetylisoniazid, NAT2 pharmacogenetics (Chen 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Isoniazid + acetylisoniazid, NAT2 pharmacogenetics (Chen 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Chen B, Shi H-Q, Feng MR, Wang X-H, Cao X-M, Cai W-M (2022).
  Population Pharmacokinetics and Pharmacodynamics of Isoniazid and its
  Metabolite Acetylisoniazid in Chinese Population. *Front Pharmacol*
  13:932686.
* Article: [doi:10.3389/fphar.2022.932686](https://doi.org/10.3389/fphar.2022.932686)

Chen 2022 built an integrated parent-metabolite population PK model for oral
isoniazid (INH) and its acetylation product acetylisoniazid (AcINH) in healthy
Chinese adults and Chinese tuberculosis patients (Figure 1 of the paper). INH is
a one-compartment model with first-order absorption (`Ka`); a fraction `FM` of
INH clearance forms AcINH, which has its own one-compartment disposition with a
first-order elimination rate constant `K30`. The remaining fraction of INH
clearance, `1 - FM`, and the AcINH elimination both leave the system.

The *N*-acetyltransferase 2 (NAT2) genotype was the only retained covariate, and
the paper reports two equally final parameterisations of it (Table 3):

* **Final Model (2)** -- a per-allele model, carried here as
  **`Chen_2022_isoniazid`**. The number of each of the NAT2 `*5`, `*6` and `*7`
  slow alleles enters exponentially on both `CL/F` and `FM`. This is the paper's
  preferred model: it gave the largest objective-function drop
  (Delta OFV = -315.5 against -275.3 for the class model).
* **Final Model (1)** -- a three-level genotype-class model, carried here as
  **`Chen_2022_isoniazid_nat2class`**. A single class score (0 = wt/wt,
  1 = m/wt, 2 = m/m) enters exponentially on `CL/F` and `FM`.

```{r load-models}
mod_allele <- rxode2::rxode2(readModelDb("Chen_2022_isoniazid"))
mod_class  <- rxode2::rxode2(readModelDb("Chen_2022_isoniazid_nat2class"))
mod_allele
```

The paper reported and plotted all concentrations in umol/L. The model files
store doses in mg and convert to umol with the isoniazid molecular weight
(137.14 g/mol), so `Cc` and `Cc_acinh` come out in umol/L to match the paper's
figures and Table 2.

# Population

The analysis pooled 202 subjects across three cohorts (Table 1 and the Results
section "INH and AcINH in Relation to NAT2 Genotypes"):

* **Study 1** -- 24 healthy adult men, single oral 300 mg INH, rich sampling over
  0-14 h.
* **Study 2** -- 21 healthy adults, single oral 320 mg INH (bioequivalence
  study).
* **Patients** -- 157 adults with pulmonary tuberculosis (89 men, 68 women; mean
  age 42.2 y, weight 56.5 kg) on 7-14 days of combination therapy, sampled 2
  and/or 6 h after a dose.

122 subjects formed the model-building (index) group and 80 the validation
group; the final model was fit to both. Eight NAT2 diplotypes were observed
(`*4/*4`, `*4/*5`, `*4/*6`, `*4/*7`, `*5/*7`, `*6/*6`, `*6/*7`, `*7/*7`), with
`*4/*4` (wild type) the most common.

```{r population}
mod_allele$meta$population[c("n_subjects", "weight_range", "disease_state")]
```

# Source trace

Every `ini()` value is traced to Table 3 of Chen 2022 in an in-file comment next
to the parameter. The table below collects the structural values of both final
models.

| Parameter | Final Model (2) (`Chen_2022_isoniazid`) | Final Model (1) (`Chen_2022_isoniazid_nat2class`) | Source |
|-----------|----------------------------------------|---------------------------------------------------|--------|
| `Ka` (1/h)         | 3.99  | 3.91  | Table 3 theta1 |
| `CL/F` (L/h), `*4/*4` | 30.2  | 28.7  | Table 3 theta2 |
| `K30` (1/h)        | 0.30  | 0.41  | Table 3 theta3 |
| `V2/F` (L)         | 51.4  | 54.1  | Table 3 theta4 |
| `V3/F` (L)         | 12.4  | 17.2  | Table 3 theta5 |
| `FM`               | 0.86  | 0.88  | Table 3 theta6 |
| `*5` effect on `CL/F` | -0.77 (theta7) | class score -0.55 (theta7) | Table 3 |
| `*6` effect on `CL/F` | -0.60 (theta8) | -- | Table 3 |
| `*7` effect on `CL/F` | -0.29 (theta9) | -- | Table 3 |
| `*5` effect on `FM`   | -0.72 (theta10) | class score -0.47 (theta10) | Table 3 |
| `*6` effect on `FM`   | -0.45 (theta11) | -- | Table 3 |
| `*7` effect on `FM`   | -0.14 (theta12) | -- | Table 3 |
| INH proportional error | 0.339 | 0.333 | Table 3 sigma INH |
| AcINH proportional error | 0.295 | 0.302 | Table 3 sigma AcINH |

The paper's Table 3 "M341 / M590 / M870" rows label the `*5` (341T>C), `*6`
(590G>A) and `*7` (857G>A) allele effects; "M803" in the Final Model (2)
footnote is a misprint of the 857 position given in the Methods "Genotyping"
section. The IIV columns are reported as `omega x 100` (a percentage), so each is
squared on the fraction scale to recover the log-scale variance.

# NAT2 covariate effects reproduce the reported fold-changes

The per-allele exponents back-transform directly to the fold-changes the
abstract and Results quote: one copy of `*5`, `*6` or `*7` leaves `CL/F` at
46.3 %, 54.9 % and 74.8 % of `*4/*4`, and `FM` at 48.7 %, 63.8 % and 86.9 %.

```{r allele-fractions}
allele_cl <- exp(c(`*5` = -0.77, `*6` = -0.60, `*7` = -0.29))
allele_fm <- exp(c(`*5` = -0.72, `*6` = -0.45, `*7` = -0.14))
reported_cl <- c(`*5` = 0.463, `*6` = 0.549, `*7` = 0.748)
reported_fm <- c(`*5` = 0.487, `*6` = 0.638, `*7` = 0.869)

data.frame(
  allele      = names(allele_cl),
  CLF_model   = round(allele_cl, 3),
  CLF_paper   = reported_cl,
  FM_model    = round(allele_fm, 3),
  FM_paper    = reported_fm
)

# These are exact algebra (no simulation), so a tight bound is correct here.
stopifnot(
  max(abs(allele_cl - reported_cl)) < 0.005,
  max(abs(allele_fm - reported_fm)) < 0.005
)
```

The class model's score coefficient of -0.55 on `CL/F` gives the m/wt and m/m
`CL/F` as `exp(-0.55) = 57.7 %` and `exp(-1.1) = 33.3 %` of wt/wt, consistent
with the Results statement that the IIV on `CL/F` fell from 56.6 % to 28.1 %
once NAT2 was added.

# Typical-value isoniazid profiles and NCA

We simulate a single 300 mg oral dose at the typical-value parameters (no random
effects) for the three NAT2 genotype classes and check the INH NCA against the
observed values in Table 2 (healthy wt/wt subjects). For the per-allele model we
use representative diplotypes: `*4/*4` (wt/wt), `*4/*6` (one `*6`, m/wt) and
`*6/*6` (two `*6`, m/m), the most common mutant alleles in the cohort.

```{r typical-profiles}
tgrid <- seq(0, 14, by = 0.1)

# Covariate sets for the per-allele model, keyed by genotype class.
allele_cov <- list(
  "wt/wt" = c(0, 0, 0),
  "m/wt"  = c(0, 1, 0),
  "m/m"   = c(0, 2, 0)
)

sim_allele_typ <- function(label, counts) {
  cov <- data.frame(
    SNP_NAT2_RS1801280_C_COUNT = counts[1],
    SNP_NAT2_RS1799930_A_COUNT = counts[2],
    SNP_NAT2_RS1799931_A_COUNT = counts[3]
  )
  dose <- data.frame(id = 1, time = 0, amt = 300, evid = 1L, cmt = "depot")
  obs <- tidyr::expand_grid(id = 1, time = tgrid, cmt = c("Cc", "Cc_acinh")) |>
    dplyr::mutate(amt = NA_real_, evid = 0L)
  ev <- dplyr::bind_rows(dose, obs) |>
    dplyr::arrange(time, cmt) |>
    cbind(cov)
  s <- rxode2::rxSolve(mod_allele, ev, omega = NA, sigma = NA,
                       returnType = "data.frame")
  s |>
    dplyr::distinct(time, .keep_all = TRUE) |>
    dplyr::arrange(time) |>
    dplyr::transmute(genotype = label, time, Cc, Cc_acinh)
}

typ <- dplyr::bind_rows(mapply(sim_allele_typ, names(allele_cov), allele_cov,
                               SIMPLIFY = FALSE))
head(typ)
```

PKNCA on the INH typical-value profiles:

```{r pknca-inh}
conc_df <- typ |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id = genotype, time, Cc) |>
  dplyr::mutate(genotype = id)

dose_df <- typ |>
  dplyr::distinct(genotype) |>
  dplyr::mutate(id = genotype, time = 0, amt = 300 / 137.14 * 1000)

conc_obj <- PKNCA::PKNCAconc(
  dplyr::select(conc_df, genotype, id, time, Cc),
  Cc ~ time | genotype + id, concu = "umol/L", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | genotype + id, doseu = "umol")

intervals <- data.frame(start = 0, end = Inf,
                        cmax = TRUE, tmax = TRUE,
                        aucinf.obs = TRUE, half.life = TRUE)
nca_inh <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
nca_wide <- as.data.frame(nca_inh$result) |>
  dplyr::select(genotype, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
nca_wide
```

Chen 2022 Table 2 reports the observed INH NCA in the healthy single-dose study.
For wt/wt subjects: `Cmax` 36.0 umol/L, `AUC` 75.5 umol*h/L, `t1/2` 1.15 h,
`CL` 30.05 L/h. The typical-value AUC matches to within a few percent (the
`AUC0-inf = Dose/CL` identity is the load-bearing structural check), and the
terminal half-life is reproduced. The typical-value `Cmax` lands below the
observed mean because the authors themselves report the absorption phase was
poorly characterised ("the absorption phase was not effectively estimated ...
inaccurate in predicting Ka and Cmax", Discussion limitations).

```{r pknca-inh-gate}
wt <- nca_wide[nca_wide$genotype == "wt/wt", ]
# AUC0-inf = Dose/CL is an identity for a linear model, so a mis-transcribed
# clearance, dose or molecular weight would move this by tens of percent.
stopifnot(abs(wt$aucinf.obs - 75.5) / 75.5 < 0.10)
# Terminal half-life from the structural parameters (ln2 * V2 / CL/F).
stopifnot(abs(wt$half.life - 1.15) / 1.15 < 0.15)
```

# Known deviation: acetylisoniazid exposure in Final Model (2)

The printed Final Model (2) metabolite parameters give an apparent AcINH
clearance of `K30 x V3/F = 0.30 x 12.4 = 3.7 L/h`. With that value the
typical-value AcINH exposure is about twice the observed NCA (Table 2 wt/wt:
`Cmax` 32.5 umol/L, `AUC` 245.4 umol*h/L), even though INH reproduces. Final
Model (1), whose `K30 x V3/F = 0.41 x 17.2 = 7.1 L/h`, reproduces the AcINH AUC
to within ~10 %. The paper's own Discussion states the AcINH clearance
"was estimated as 11.3 L*h^-1", which neither printed `K30 x V3/F` product
equals; the AcINH disposition parameters are internally inconsistent as printed.

Both models encode exactly what Table 3 reports -- no value is tuned. The INH
layer, which is what drives the exposure-response / target-attainment analysis
in the paper, is sound in both.

```{r acinh-deviation}
acinh_auc <- function(df) {
  d <- dplyr::filter(df, genotype == "wt/wt")
  sum(diff(d$time) * (head(d$Cc_acinh, -1) + tail(d$Cc_acinh, -1)) / 2)
}
sim_class_typ <- function(label, rapid, slow) {
  cov <- data.frame(NAT2_RAPID = rapid, NAT2_SLOW = slow)
  dose <- data.frame(id = 1, time = 0, amt = 300, evid = 1L, cmt = "depot")
  obs <- tidyr::expand_grid(id = 1, time = tgrid, cmt = c("Cc", "Cc_acinh")) |>
    dplyr::mutate(amt = NA_real_, evid = 0L)
  ev <- dplyr::bind_rows(dose, obs) |> dplyr::arrange(time, cmt) |> cbind(cov)
  s <- rxode2::rxSolve(mod_class, ev, omega = NA, sigma = NA,
                       returnType = "data.frame")
  s |> dplyr::distinct(time, .keep_all = TRUE) |> dplyr::arrange(time) |>
    dplyr::transmute(genotype = label, time, Cc, Cc_acinh)
}
typ_class <- dplyr::bind_rows(
  sim_class_typ("wt/wt", 1, 0),
  sim_class_typ("m/wt", 0, 0),
  sim_class_typ("m/m", 0, 1)
)

auc_allele <- acinh_auc(typ)        # Final Model (2)
auc_class  <- acinh_auc(typ_class)  # Final Model (1)
data.frame(
  model = c("Final Model (2) per-allele", "Final Model (1) class"),
  acinh_auc0_14 = round(c(auc_allele, auc_class), 0),
  observed = 245.4
)

# Final Model (1) reproduces AcINH AUC; Final Model (2) is the ~2x deviation.
# These bounds document the known behaviour; they fail only if a structural
# value (K30, V3/F, FM or the molecular weight) is mis-transcribed.
stopifnot(abs(auc_class - 245.4) / 245.4 < 0.20)
stopifnot(auc_allele / auc_class > 1.5)
```

# Replicating the concentration-time profiles (Figure 4)

Typical-value INH and AcINH profiles by NAT2 genotype class reproduce the shape
of Chen 2022 Figure 4: slower INH elimination (and a higher, flatter INH peak)
with increasing numbers of slow alleles, and the opposite ordering for AcINH as
less INH is acetylated.

```{r figure4, fig.alt = "Typical-value INH and AcINH concentration-time profiles by NAT2 genotype class"}
plt <- typ |>
  tidyr::pivot_longer(c(Cc, Cc_acinh), names_to = "analyte", values_to = "conc") |>
  dplyr::mutate(analyte = dplyr::recode(analyte, Cc = "INH", Cc_acinh = "AcINH"),
                analyte = factor(analyte, c("INH", "AcINH")),
                genotype = factor(genotype, c("wt/wt", "m/wt", "m/m")))

ggplot(plt, aes(time, conc, colour = genotype)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~analyte, scales = "free_y") +
  labs(x = "Time (h)", y = "Concentration (umol/L)", colour = "NAT2") +
  theme_bw()
```

# A small stochastic cohort

A 100-subject-per-class stochastic simulation (the full between-subject
variability of Final Model (2)) illustrates the spread the VPC in Figure 4
summarises. Residual error is left off so the curves show the structural plus
IIV spread.

```{r vpc, fig.alt = "Stochastic INH concentration-time profiles by NAT2 genotype class"}
rxode2::rxSetSeed(1234)
nsub <- 100
sim_cohort <- function(label, counts) {
  cov <- data.frame(
    SNP_NAT2_RS1801280_C_COUNT = counts[1],
    SNP_NAT2_RS1799930_A_COUNT = counts[2],
    SNP_NAT2_RS1799931_A_COUNT = counts[3]
  )
  dose <- tidyr::expand_grid(id = seq_len(nsub), time = 0) |>
    dplyr::mutate(amt = 300, evid = 1L, cmt = "depot")
  obs <- tidyr::expand_grid(id = seq_len(nsub), time = seq(0, 14, by = 0.5),
                            cmt = "Cc") |>
    dplyr::mutate(amt = NA_real_, evid = 0L)
  ev <- dplyr::bind_rows(dose, obs) |>
    dplyr::arrange(id, time) |>
    cbind(cov)
  s <- rxode2::rxSolve(mod_allele, ev, returnType = "data.frame")
  s |> dplyr::mutate(genotype = label)
}
cohort <- dplyr::bind_rows(mapply(sim_cohort, names(allele_cov), allele_cov,
                                  SIMPLIFY = FALSE))

qs <- cohort |>
  dplyr::group_by(genotype = factor(genotype, c("wt/wt", "m/wt", "m/m")), time) |>
  dplyr::summarise(med = median(Cc),
                   lo = quantile(Cc, 0.05),
                   hi = quantile(Cc, 0.95), .groups = "drop")

ggplot(qs, aes(time, med, colour = genotype, fill = genotype)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time (h)", y = "INH concentration (umol/L)",
       colour = "NAT2", fill = "NAT2") +
  theme_bw()
```

# Assumptions and deviations

* **Units.** The paper plots and tabulates concentrations in umol/L. Doses are
  carried in mg and converted with the isoniazid molecular weight 137.14 g/mol,
  which is consistent with the paper's own unit pairs (e.g. the calibration
  range 15.89 mg/L = 115.9 umol/L, and 19.7 ug*h/mL = 143.4 umol*h/L).
* **`FM` is not bounded above by 1.** Chen 2022 uses an exponential covariate
  model `FM = theta6 * exp(sum  thetak * allele)` (Eq. 1 / Table 3 footnote), so the
  wt/wt `FM` of 0.86-0.88 is a free parameter rather than a logit-constrained
  fraction. The model encodes it as printed.
* **Acetylisoniazid exposure, Final Model (2).** As documented above, the
  printed Final Model (2) `K30 x V3/F` gives an AcINH exposure about twice the
  observed NCA, while Final Model (1) reproduces it. Both are faithful
  transcriptions of Table 3; nothing is tuned.
* **Isoniazid Cmax.** The typical-value `Cmax` sits below the observed mean
  because the authors report the absorption phase (and hence `Ka` / `Cmax`) was
  poorly estimated from the sparse early sampling. The AUC and terminal
  half-life, which depend on `CL/F` and `V2/F`, are reproduced.
* **Genotype representatives.** Table 2's NCA is reported by genotype class
  (wt/wt, m/wt, m/m). For the per-allele model these classes are illustrated
  with the most common diplotypes (`*4/*4`, `*4/*6`, `*6/*6`); other mutant
  alleles give the per-allele fold-changes verified above.
* **No weight or renal covariate.** Only NAT2 was retained in the final model
  (Results), so body weight, age, creatinine clearance and the transaminases
  that were screened do not enter either model.
```
