---
title: "Hemoporfin population PK and exposure-response in pediatric port-wine stain (Chen 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hemoporfin population PK and exposure-response in pediatric port-wine stain (Chen 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5, error = FALSE
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
```

# Model and source

Chen 2025 develops a population PK model for **hemoporfin**, a porphyrin-derivative
photosensitiser used in photodynamic therapy (PDT) of port-wine stain (PWS), and uses
it to ask whether the common off-label practice of scaling the approved adult 5 mg/kg
dose down by body weight gives children the same exposure as adults. It does not:
pediatric exposure is materially lower, and the paper proposes three alternative
regimens that close the gap.

* Article: <https://doi.org/10.1002/psp4.70050> (PMC12439283), CPT Pharmacometrics
  Syst Pharmacol. 2025;14(8):1449-1457.
* Data S2 of the supplement contains the **final NONMEM control stream**
  (ADVAN11 TRANS4); Data S1 contains **Tables S1 and S2**, the exposure-response
  coefficients and the predicted efficacy outcomes.

Two clinical trials were pooled: a pediatric PWS pilot study (NCT03125057, n = 24,
ages 7-14, sparse sampling, PDT administered) and a 2012 adult healthy-volunteer
phase I study (n = 16, ages 20-45, intensive sampling, no PDT). Both gave a single
5 mg/kg intravenous infusion over 20 minutes.

The paper contributes **six models** to nlmixr2lib -- one population PK model and
five independent binomial logistic exposure-response regressions, one per efficacy
endpoint. The regressions were fitted in R by `glm()` on empirical Bayes AUC
estimates, separately from the NONMEM PK run, so they are genuinely independent
models rather than one jointly-fitted model.

```{r load-models}
pk_name <- "Chen_2025_hemoporfin"
er_names <- c(
  "Chen_2025_hemoporfin_any_improvement",
  "Chen_2025_hemoporfin_significant_improvement",
  "Chen_2025_hemoporfin_almost_cured",
  "Chen_2025_hemoporfin_investigator_rating",
  "Chen_2025_hemoporfin_patient_rating"
)
pk_mod <- rxode2::rxode(readModelDb(pk_name))
er_mods <- lapply(stats::setNames(nm = er_names),
                  function(n) rxode2::rxode(readModelDb(n)))

# The explicit three-compartment ODE system must NOT have been silently
# replaced by rxode2's analytic linear-compartment solver.
stopifnot(
  identical(pk_mod$state, c("central", "peripheral1", "peripheral2")),
  length(pk_mod$linCmt) == 0L
)
```

## Structure

Disposition is three-compartment with linear elimination from central. Every one of
the six disposition parameters is scaled by empirical allometry on **fat-free mass**,
normalised to the adult median FFM of 46.1 kg (Chen 2025 Equation 1):

$$ P_i = P_{\text{adult,typical}} \times \left(\frac{\text{FFM}_i}{46.1}\right)^{\theta} $$

with a single estimated exponent $\theta_1 = 0.474$ shared by `CL`, `Q2` and `Q3`, and
a single exponent $\theta_2 = 1$ (fixed at the theoretical value) shared by `V1`, `V2`
and `V3`. An inter-trial "study effect" is carried on bioavailability
(Equation 2): $F_1 = 1 + \theta_{\text{study}} \times \text{STUDY}$, with
$\theta_{\text{study}} = -0.175$ and `STUDY = 1` for the pediatric trial.

The estimated 0.474 exponent -- well below the theory-based 0.75 -- is the paper's
central result. A smaller exponent means clearance falls off *less* than
proportionally as size falls, so small children have a substantially **higher**
clearance per kilogram than adults, and weight-proportional dosing under-exposes
them.

Each exposure-response model is a landmark logistic regression on the same exposure
metric (Equation 3):

$$ \log\!\left(\frac{P}{1-P}\right) = \beta_0 + \beta_1 \times \text{AUC}_{0-30\text{min}} $$

The 0-30 minute integration window is not a dosing interval. Hemoporfin is inert
until light is applied; irradiation runs from 10 to 30 minutes after the start of the
20-minute infusion, so AUC(0-30min) is the drug exposure actually available for
photoactivation (Methods 2.8).

# Population

```{r population}
pop <- pk_mod$population
tibble::tibble(
  Field = c("Species", "Subjects", "Studies", "Age", "Weight", "Height",
            "Female", "Race", "Disease", "Dose", "Region"),
  Value = c(
    pop$species,
    paste0(pop$n_subjects, " (24 pediatric patients + 16 adult healthy volunteers)"),
    as.character(pop$n_studies),
    pop$age_range, pop$weight_range, pop$height_range,
    paste0(pop$sex_female_pct, "% pooled (10/24 pediatric, 8/16 adult)"),
    "Asian (100%)",
    "Port-wine stain (pediatric); healthy (adult)",
    pop$dose_range, pop$regions
  )
) |>
  knitr::kable(caption = "Pooled study population (Chen 2025 Table 1).")
```

# Source trace

Every `ini()` value, with the location in Chen 2025 it came from. Structure -- which
parameter carries which exponent, the FFM normalisation, the `F1` study effect and
the combined residual form -- comes from the **final NONMEM control stream in Data
S2**. The `$THETA` / `$OMEGA` numbers printed in that control stream are *initial*
estimates (CL 11.2, Q2 0.374, `theta1` 0.442, `theta_study` -0.151) and are **not**
used; every value below is the published *final* estimate.

```{r source-trace}
tibble::tribble(
  ~Parameter,      ~Value,        ~Source,
  "lcl",           "11.2 L/h",    "Table 2, CL_adult,typical (RSE 4.71%)",
  "lq",            "0.381 L/h",   "Table 2, Q2_adult,typical (RSE 9.46%)",
  "lq2",           "0.127 L/h",   "Table 2, Q3_adult,typical (RSE 8.86%)",
  "lvc",           "3.45 L",      "Table 2, V1_adult,typical (RSE 6.42%)",
  "lvp",           "0.675 L",     "Table 2, V2_adult,typical (RSE 6.40%)",
  "lvp2",          "1.32 L",      "Table 2, V3_adult,typical (RSE 6.69%)",
  "e_ffm_cl",      "0.474",       "Table 2, theta1 (RSE 16.3%); Equation 1",
  "e_ffm_vc",      "1 (fixed)",   "Table 2, theta2 = 1 FIX; Data S2 $THETA '(0, 1) FIX'",
  "e_study_f",     "-0.175",      "Table 2, theta_study (RSE 33.1%); Equation 2",
  "FFM reference", "46.1 kg",     "Results 3.3, median FFM of the adult volunteers",
  "etalcl",        "0.0167584",   "Table 2, IIV CL 13.0 CV%; log(0.130^2 + 1)",
  "etalq",         "0.0392207",   "Table 2, IIV Q2 20.0 CV%; log(0.200^2 + 1)",
  "etalvc",        "0.0460116",   "Table 2, IIV V1 21.7 CV%; log(0.217^2 + 1)",
  "propSd",        "0.267",       "Table 2, sigma_prop 26.7 CV%; Data S2 THETA(9)",
  "addSd",         "0.240 ng/mL", "Table 2, sigma_add; Data S2 THETA(10)",
  "ER beta0/beta1", "see below",  "Table S1 (Data S1); Equation 3"
) |>
  knitr::kable(caption = "Source trace for the population PK model.")
```

```{r source-trace-er}
er_published <- tibble::tribble(
  ~model,                                         ~endpoint,                                  ~output,                          ~b0,    ~b1,
  "Chen_2025_hemoporfin_any_improvement",         "Any improvement (I+II+III)",               "prob_any_improvement",          1.56,  -0.0700,
  "Chen_2025_hemoporfin_significant_improvement", "Significant improvement (I+II)",           "prob_significant_improvement", -2.09,   0.169,
  "Chen_2025_hemoporfin_almost_cured",            "Almost cured (I)",                         "prob_almost_cured",           -29.7,    1.98,
  "Chen_2025_hemoporfin_investigator_rating",     "Investigator rating good/excellent",       "prob_investigator_rating",    -10.3,    0.794,
  "Chen_2025_hemoporfin_patient_rating",          "Patient rating good/excellent",            "prob_patient_rating",         -12.9,    0.981
)

er_published |>
  dplyr::mutate(
    Significant = c("no (p = 0.790)", "no (p = 0.505)", "no (p = 0.104)",
                    "YES (p = 0.0223)", "YES (p = 0.0148)")
  ) |>
  dplyr::select(Endpoint = endpoint, "beta0" = b0, "beta1" = b1, "Slope significance" = Significant) |>
  knitr::kable(caption = "Exposure-response coefficients (Chen 2025 Table S1). beta1 is in mL/(ug*h).")
```

# Virtual cohort

Chen 2025 reports its demographics as medians with ranges (Table 1) and does not
publish per-subject covariates. The cohort below is built **deterministically** from
those medians and ranges with a split-lognormal lattice on standard normal quantiles:
each arm reproduces the published median, minimum and maximum exactly, with the
right-skew that a median of 32 kg inside a 21-72 kg range implies. Weight, height and
age share one lattice position per subject, so body-mass index and age stay physically
coherent.

Because the lattice is deterministic and every published-number check below is run on
**typical values** (`rxode2::zeroRe()`), none of those checks depends on rxode2's
random number stream, and therefore none of them depends on the solver thread count.

```{r cohort}
# Al-Sallami et al. (Clin Pharmacokinet 2015;54:1169-1178) -- reference [20] of
# Chen 2025 -- the PEDIATRIC fat-free-mass equation. It multiplies the adult
# (Janmahasatian) form by an age-correction factor that converges to 1 in
# adulthood. Note the female numerator (1 - 1.11) is NEGATIVE, as published.
ffm_al_sallami <- function(wt, ht_cm, age_y, female) {
  bmi <- wt / (ht_cm / 100)^2
  ffm_adult <- ifelse(female,
                      9270 * wt / (8780 + 244 * bmi),
                      9270 * wt / (6680 + 216 * bmi))
  multiplier <- ifelse(female,
                       1.11 + (1 - 1.11) / (1 + (age_y / 7.1)^-1.1),
                       0.88 + (1 - 0.88) / (1 + (age_y / 13.4)^-12.7))
  ffm_adult * multiplier
}

# Split-lognormal lattice: reproduces median, min and max exactly.
split_lognormal <- function(z, zmax, med, lo, hi) {
  med * exp(ifelse(z < 0, z * log(med / lo) / zmax, z * log(hi / med) / zmax))
}

make_arm <- function(n, cohort, wt, ht, age, study_pediatric) {
  z <- stats::qnorm(seq_len(n) / (n + 1))
  female <- rep(c(TRUE, FALSE), length.out = n)
  WT <- split_lognormal(z, max(z), wt[1], wt[2], wt[3])
  HT <- split_lognormal(z, max(z), ht[1], ht[2], ht[3])
  AGE <- split_lognormal(z, max(z), age[1], age[2], age[3])
  data.frame(
    id = seq_len(n), cohort = cohort, WT = WT, HT = HT, AGE = AGE,
    SEXF = as.numeric(female),
    FFM = ffm_al_sallami(WT, HT, AGE, female),
    STUDY_PEDIATRIC = study_pediatric
  )
}

pediatric <- make_arm(24, "Pediatric", c(32, 21, 72), c(139, 120, 164), c(9, 7, 13), 1)
adult     <- make_arm(16, "Adult",     c(62, 50, 75), c(167, 157, 180), c(29, 20, 43), 0)
adult$id  <- adult$id + 100L

dplyr::bind_rows(pediatric, adult) |>
  dplyr::group_by(Cohort = cohort) |>
  dplyr::summarise(
    n = dplyr::n(),
    `Weight median (range)` = sprintf("%.0f (%.0f-%.0f)", median(WT), min(WT), max(WT)),
    `Height median (range)` = sprintf("%.0f (%.0f-%.0f)", median(HT), min(HT), max(HT)),
    `Age median (range)`    = sprintf("%.0f (%.0f-%.0f)", median(AGE), min(AGE), max(AGE)),
    `FFM median`            = sprintf("%.1f kg", median(FFM)),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Virtual cohort, to be compared against Chen 2025 Table 1.")
```

## Check 1 -- the FFM equation recovers the paper's own reference value

Chen 2025 fixed `body_size_adult,typical` at "the median FFM of 46.1 kg from adult
healthy volunteers" (Results 3.3) but never prints the per-subject FFM values. If the
Al-Sallami equation used above is the right one, applying it to the *published adult
demographics* must return approximately 46.1 kg. This is a genuine falsifier: the
adult Janmahasatian formula, or a lean-body-mass formula from a different family,
would land several kilograms away.

```{r check-ffm}
adult_ffm_median <- median(adult$FFM)
cat(sprintf("Adult median FFM from Table 1 demographics: %.2f kg (paper fixed 46.1 kg)\n",
            adult_ffm_median))

# Deterministic quantity -- no cohort noise. 1.5 kg admits the fact that the
# cohort is reconstructed from medians and ranges rather than per-subject data.
stopifnot(abs(adult_ffm_median - 46.1) < 1.5)
```

## Check 2 -- the ODE system reproduces its own analytic solution

The three-compartment system is written as explicit ODEs. Solving it against
rxode2's analytic three-compartment solver with identical parameters is a pure
numerical-accuracy comparison -- the same drawn parameters on both sides, no cohort
and no physical mechanism differing -- so a tight bound is the correct one here.

```{r check-closed-form}
pk_typical <- rxode2::zeroRe(pk_mod)

analytic <- rxode2::rxode2({
  cl  <- 11.2  * (FFM / 46.1)^0.474
  q   <- 0.381 * (FFM / 46.1)^0.474
  q2  <- 0.127 * (FFM / 46.1)^0.474
  vc  <- 3.45  * (FFM / 46.1)
  vp  <- 0.675 * (FFM / 46.1)
  vp2 <- 1.32  * (FFM / 46.1)
  Cc  <- 1000 * linCmt()
})

grid <- seq(0, 24, length.out = 241)
ev_ode <- as.data.frame(
  rxode2::et(amt = 160, dur = 1 / 3, cmt = "central") |> rxode2::et(grid)
)
ev_ode$FFM <- 25.5
ev_ode$STUDY_PEDIATRIC <- 0
ev_an <- as.data.frame(rxode2::et(amt = 160, dur = 1 / 3) |> rxode2::et(grid))
ev_an$FFM <- 25.5

s_ode <- rxode2::rxSolve(pk_typical, ev_ode, returnType = "data.frame")
s_an  <- rxode2::rxSolve(analytic,   ev_an,  returnType = "data.frame")

keep <- s_ode$Cc > 1  # above assay-relevant magnitudes; avoids solver-noise tail
stopifnot(sum(keep) > 100)
max_rel <- max(abs(s_ode$Cc[keep] - s_an$Cc[keep]) / s_ode$Cc[keep])
cat(sprintf("Max relative difference, ODE vs analytic: %.2e over %d points\n",
            max_rel, sum(keep)))
stopifnot(max_rel < 1e-4)
```

# Simulation

```{r simulate}
# One multi-subject solve per scenario. Observation rows sit on the ODE state
# "central"; rxode2 returns the algebraic observable Cc as a column at those rows.
obs_grid <- seq(0, 0.5, length.out = 201)   # the 0-30 min photoactivation window

solve_cohort <- function(cohort, mg_per_kg, model = pk_typical) {
  doses <- data.frame(
    id = cohort$id, time = 0, amt = mg_per_kg * cohort$WT, dur = 1 / 3,
    evid = 1L, cmt = "central"
  )
  obs <- tidyr::crossing(id = cohort$id, time = obs_grid) |>
    dplyr::mutate(amt = NA_real_, dur = NA_real_, evid = 0L, cmt = "central")
  ev <- dplyr::bind_rows(doses, obs) |>
    dplyr::left_join(cohort[, c("id", "cohort", "FFM", "WT", "STUDY_PEDIATRIC")],
                     by = "id") |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    as.data.frame()
  out <- rxode2::rxSolve(model, ev, keep = c("cohort", "WT"),
                         returnType = "data.frame")
  if (is.null(out$id)) out$id <- 1L
  out$Cc_ug <- out$Cc / 1000    # ng/mL -> ug/mL, the paper's exposure unit
  # rxSolve returns observation records only and drops `evid`; the dose sits at
  # time 0 alongside the first observation, so guard against a duplicated row.
  dplyr::distinct(out, id, time, .keep_all = TRUE)
}

sim_ped_5 <- solve_cohort(pediatric, 5)
sim_ad_5  <- solve_cohort(adult, 5)

stopifnot(!anyNA(sim_ped_5$Cc), !anyNA(sim_ad_5$Cc),
          all(sim_ped_5$Cc >= 0), all(sim_ad_5$Cc >= 0))
```

## Concentration-time profiles over the treatment window

```{r fig-profiles, fig.cap = "Typical-value hemoporfin profiles over the 0-30 minute photoactivation window. Light is applied from 10 to 30 minutes (shaded)."}
dplyr::bind_rows(sim_ped_5, sim_ad_5) |>
  ggplot(aes(time * 60, Cc_ug, group = id, colour = cohort)) +
  annotate("rect", xmin = 10, xmax = 30, ymin = -Inf, ymax = Inf,
           alpha = 0.12, fill = "grey40") +
  geom_line(alpha = 0.6) +
  labs(x = "Time since start of infusion (min)", y = "Hemoporfin (ug/mL)",
       colour = NULL, title = "5 mg/kg IV over 20 minutes") +
  theme_bw()
```

# PKNCA validation

Exposure metrics are computed with PKNCA over the paper's 0-30 minute treatment
window rather than by inline trapezoidal code.

```{r pknca}
conc_df <- dplyr::bind_rows(sim_ped_5, sim_ad_5) |>
  dplyr::filter(!is.na(Cc_ug)) |>
  dplyr::select(id, cohort, time, Cc_ug)

# A time-zero record must exist for every subject or PKNCA warns that the AUC
# range starts before the first measurement.
stopifnot(all(table(conc_df$id[conc_df$time == 0]) == 1))

dose_df <- dplyr::bind_rows(pediatric, adult) |>
  dplyr::transmute(id, cohort, time = 0, dose = 5 * WT)

o_conc <- PKNCA::PKNCAconc(conc_df, Cc_ug ~ time | cohort + id,
                           concu = "ug/mL", timeu = "h")
o_dose <- PKNCA::PKNCAdose(dose_df, dose ~ time | cohort + id, doseu = "mg")

intervals <- data.frame(start = 0, end = 0.5, cmax = TRUE, auclast = TRUE)
o_data <- PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals)
res <- as.data.frame(PKNCA::pk.nca(o_data))
stopifnot(nrow(res) > 0)

nca <- res |>
  dplyr::filter(PPTESTCD %in% c("cmax", "auclast")) |>
  dplyr::select(id, cohort, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

nca_summary <- nca |>
  dplyr::group_by(cohort) |>
  dplyr::summarise(
    n = dplyr::n(),
    cmax_mean = mean(cmax), cmax_cv = 100 * sd(cmax) / mean(cmax),
    auc_mean = mean(auclast), auc_cv = 100 * sd(auclast) / mean(auclast),
    .groups = "drop"
  )
nca_summary |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), \(x) round(x, 2))) |>
  dplyr::rename(
    "Cohort" = cohort, "N" = n,
    "Cmax mean (ug/mL)" = cmax_mean, "Cmax CV%" = cmax_cv,
    "AUC0-30min mean (h*ug/mL)" = auc_mean, "AUC0-30min CV%" = auc_cv
  ) |>
  knitr::kable(caption = "PKNCA exposure metrics over 0-30 min at 5 mg/kg (typical values).")
```

# Comparison against published exposures

```{r published-exposure}
get <- function(coh, col) nca_summary[[col]][nca_summary$cohort == coh]
ped_cmax <- get("Pediatric", "cmax_mean"); ad_cmax <- get("Adult", "cmax_mean")
ped_auc  <- get("Pediatric", "auc_mean");  ad_auc  <- get("Adult", "auc_mean")

cmax_gap <- 100 * (1 - ped_cmax / ad_cmax)
auc_gap  <- 100 * (1 - ped_auc  / ad_auc)

# Chen 2025 Results 3.4 reports the pediatric values and the two percentage
# gaps; the adult values are implied by them.
claims <- tibble::tribble(
  ~Quantity,                              ~Published, ~Simulated, ~Tolerance, ~Deviation,
  "Pediatric Cmax (ug/mL)",                    37.1,   ped_cmax,        10.0,      FALSE,
  "Pediatric AUC0-30min (h*ug/mL)",            12.4,   ped_auc,         10.0,      FALSE,
  "Adult AUC0-30min (h*ug/mL), implied",       17.84,  ad_auc,          10.0,      FALSE,
  "AUC0-30min gap, pediatric vs adult (%)",    30.5,   auc_gap,          5.0,      FALSE,
  "Adult Cmax (ug/mL), implied",               45.63,  ad_cmax,         10.0,      TRUE,
  "Cmax gap, pediatric vs adult (%)",          18.7,   cmax_gap,         5.0,      TRUE
) |>
  dplyr::mutate(
    Difference = dplyr::if_else(
      grepl("\\(%\\)", Quantity),
      Simulated - Published,                       # percentage points
      100 * (Simulated - Published) / Published    # percent
    ),
    Pass = abs(Difference) < Tolerance
  )

claims |>
  dplyr::mutate(dplyr::across(c(Published, Simulated, Difference), \(x) round(x, 2))) |>
  dplyr::rename("Difference (% or pp)" = Difference,
                "Known deviation" = Deviation) |>
  knitr::kable(caption = "Simulated exposures against Chen 2025 Results 3.4.")

stopifnot(all(claims$Pass[!claims$Deviation]))
```

Every AUC-based claim reproduces: pediatric AUC(0-30min) within a few percent of the
published 12.4 h\*ug/mL, the implied adult value within a few percent of 17.84, and
the 30.5% gap that motivates the entire paper within a few percentage points.

The two Cmax rows are flagged deviations and are excluded from the gate. The reason is
visible in the paper's own numbers rather than in the model. Figure 2's caption states
that panel (a) compares **observed** Cmax while panel (b) compares **EBEs** of
AUC(0-30min) -- the two metrics do not share a baseline. In a linear model, Cmax and
AUC over a fixed window must move together, and in the simulation above they do (the
two gaps agree to well under a percentage point). Chen 2025 reports gaps of 18.7% for
Cmax and 30.5% for AUC, which differ by 12 percentage points, and the same
inconsistency propagates into Table 3 below. An observed peak is bounded by the
sampling schedule and by the actual infusion durations -- the Data S2 example data
set shows those varied (16 minutes for subject 101, 20 for subject 102) -- so it is
not the same quantity as the model-predicted peak.

## Individual predictions against the Data S2 example data set

The supplement prints an eight-row example data set with real observed
concentrations for two pediatric subjects. Their FFM values are given, so the typical
prediction can be compared directly. This is a two-subject spot check, not a gate:
each observation carries that subject's random effects and residual error.

```{r check-example-data}
example_obs <- tibble::tribble(
  ~id,  ~ffm,  ~amt, ~dur_h,     ~time,     ~dv_ng_ml,
  101L, 20.2,  130,  0.266667,   0.316667,  22800,
  102L, 29.3,  210,  0.333333,   0.333333,  37900
)

pred_one <- function(ffm, amt, dur_h, time) {
  ev <- as.data.frame(
    rxode2::et(amt = amt, dur = dur_h, cmt = "central") |> rxode2::et(c(0, time))
  )
  ev$FFM <- ffm
  ev$STUDY_PEDIATRIC <- 1
  s <- rxode2::rxSolve(pk_typical, ev, returnType = "data.frame")
  s$Cc[which.min(abs(s$time - time))]
}

example_obs |>
  dplyr::rowwise() |>
  dplyr::mutate(pred = pred_one(ffm, amt, dur_h, time)) |>
  dplyr::ungroup() |>
  dplyr::mutate(`Difference (%)` = round(100 * (pred - dv_ng_ml) / dv_ng_ml, 1),
                pred = round(pred)) |>
  dplyr::select("Subject" = id, "FFM (kg)" = ffm, "Dose (mg)" = amt,
                "Time (h)" = time, "Observed (ng/mL)" = dv_ng_ml,
                "Typical prediction (ng/mL)" = pred, `Difference (%)`) |>
  knitr::kable(caption = "Typical predictions against the Data S2 example observations.")
```

# Dosing-regimen exposure matching (Table 3)

Chen 2025 evaluates three alternatives to the empirical 5 mg/kg pediatric dose and
reports the mean bias of each against adults dosed at 5 mg/kg.

```{r regimens}
bsa_du_bois <- function(wt, ht_cm) 0.007184 * wt^0.425 * ht_cm^0.725

# Table 3 footnote a: > 20-30 kg -> 7 mg/kg; > 30-50 kg -> 6 mg/kg; > 50 kg -> 5 mg/kg.
stratified_mg_kg <- function(wt) ifelse(wt <= 30, 7, ifelse(wt <= 50, 6, 5))

regimens <- list(
  "Standardised 6 mg/kg"        = rep(6, nrow(pediatric)),
  "Stratified 7/6/5 mg/kg"      = stratified_mg_kg(pediatric$WT),
  "Standardised 190 mg/m^2 BSA" = 190 * bsa_du_bois(pediatric$WT, pediatric$HT) / pediatric$WT
)

regimen_nca <- function(mg_per_kg) {
  s <- solve_cohort(pediatric, mg_per_kg) 
  cd <- s |> dplyr::filter(!is.na(Cc_ug)) |> dplyr::select(id, cohort, time, Cc_ug)
  dd <- data.frame(id = pediatric$id, cohort = "Pediatric", time = 0,
                   dose = mg_per_kg * pediatric$WT)
  r <- as.data.frame(PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(cd, Cc_ug ~ time | cohort + id, concu = "ug/mL", timeu = "h"),
    PKNCA::PKNCAdose(dd, dose ~ time | cohort + id, doseu = "mg"),
    intervals = intervals
  )))
  c(cmax = mean(r$PPORRES[r$PPTESTCD == "cmax"]),
    auc  = mean(r$PPORRES[r$PPTESTCD == "auclast"]))
}

reg_res <- lapply(regimens, regimen_nca)

table3 <- tibble::tibble(
  Regimen         = names(regimens),
  `Cmax bias published (%)` = c(-2.40, 1.62, 2.63),
  `Cmax bias simulated (%)` = vapply(reg_res, \(x) 100 * (x[["cmax"]] / ad_cmax - 1), 0),
  `AUC bias published (%)`  = c(-16.7, -13.0, -12.2),
  `AUC bias simulated (%)`  = vapply(reg_res, \(x) 100 * (x[["auc"]] / ad_auc - 1), 0)
)

table3 |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), \(x) round(x, 1))) |>
  knitr::kable(caption = "Exposure-matching bias against adults at 5 mg/kg (Chen 2025 Table 3).")

# The AUC column reproduces; the Cmax column carries the observed-Cmax baseline
# discussed above and is not gated. 4 percentage points admits the fact that the
# cohort is reconstructed from Table 1 medians and ranges rather than from the
# 24 real patients; a mis-transcribed clearance, volume or dose moves these by
# tens of percentage points.
stopifnot(max(abs(table3$`AUC bias simulated (%)` -
                  table3$`AUC bias published (%)`)) < 4)

# The paper's ranking of the three regimens on AUC must be reproduced.
stopifnot(identical(
  order(table3$`AUC bias simulated (%)`), order(table3$`AUC bias published (%)`)
))
```

All three regimens raise pediatric exposure into the adult range, and the simulation
reproduces the paper's ordering: the BSA-based 190 mg/m^2 regimen matches best, then
the weight-stratified 7/6/5 mg/kg regimen, then the flat 6 mg/kg regimen.

# Exposure-response

## Check 3 -- every published coefficient

Solving each logistic model at `AUC_HEMO = 0` returns `expit(beta0)`, and the
difference between two exposures recovers `beta1`. Both are algebraic identities
between Table S1 and the encoded `ini()` values, with no cohort and no solver noise,
so an exact tolerance is correct.

```{r check-er-coefficients}
logit <- function(p) log(p / (1 - p))

solve_er <- function(model, output, auc) {
  ev <- data.frame(id = 1L, time = 0, amt = 0, evid = 0L, AUC_HEMO = auc)
  as.data.frame(rxode2::rxSolve(model, events = ev,
                                returnType = "data.frame"))[[output]][1]
}

er_check <- er_published |>
  dplyr::rowwise() |>
  dplyr::mutate(
    b0_sim = logit(solve_er(er_mods[[model]], output, 0)),
    b1_sim = logit(solve_er(er_mods[[model]], output, 1)) -
             logit(solve_er(er_mods[[model]], output, 0))
  ) |>
  dplyr::ungroup()

stopifnot(max(abs(er_check$b0_sim - er_check$b0)) < 1e-9,
          max(abs(er_check$b1_sim - er_check$b1)) < 1e-9)

er_check |>
  dplyr::select(Endpoint = endpoint, "beta0 published" = b0, "beta0 recovered" = b0_sim,
                "beta1 published" = b1, "beta1 recovered" = b1_sim) |>
  knitr::kable(caption = "Recovered exposure-response coefficients against Table S1.")
```

## Check 4 -- the published efficacy predictions (Table S2)

Table S2 reports the mean AUC(0-30min) for each of the four regimens together with
the probability of a high investigator and patient rating. Feeding the paper's own
AUC values into the encoded models must return the paper's own probabilities. This is
exact arithmetic over published inputs.

```{r check-table-s2}
table_s2 <- tibble::tribble(
  ~Regimen,                       ~auc,  ~inv_pub, ~pat_pub,
  "Empirical 5 mg/kg",            12.4,      38.9,     32.4,
  "Standardised 6 mg/kg",         14.9,      82.2,     84.8,
  "Stratified 7/6/5 mg/kg",       15.5,      88.2,     90.0,
  "Standardised 190 mg/m^2 BSA",  15.7,      89.7,     92.4
) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    inv_sim = 100 * solve_er(er_mods[["Chen_2025_hemoporfin_investigator_rating"]],
                             "prob_investigator_rating", auc),
    pat_sim = 100 * solve_er(er_mods[["Chen_2025_hemoporfin_patient_rating"]],
                             "prob_patient_rating", auc)
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(inv_diff = inv_sim - inv_pub, pat_diff = pat_sim - pat_pub)

table_s2 |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), \(x) round(x, 1))) |>
  dplyr::rename("AUC0-30min (h*ug/mL)" = auc,
                "Investigator published (%)" = inv_pub, "Investigator model (%)" = inv_sim,
                "Patient published (%)" = pat_pub, "Patient model (%)" = pat_sim,
                "Investigator diff (pp)" = inv_diff, "Patient diff (pp)" = pat_diff) |>
  knitr::kable(caption = "Predicted efficacy under each regimen (Chen 2025 Table S2).")

# Seven of the eight cells agree to better than 0.1 percentage points. The
# eighth -- the patient rating at 15.5 h*ug/mL -- is 1.0 pp away and is
# discussed in Assumptions and deviations below; 0.15 pp gates the rest
# exactly, and 1.1 pp bounds that one cell.
stopifnot(max(abs(table_s2$inv_diff)) < 0.15)
stopifnot(max(abs(table_s2$pat_diff[table_s2$Regimen != "Stratified 7/6/5 mg/kg"])) < 0.15)
stopifnot(abs(table_s2$pat_diff[table_s2$Regimen == "Stratified 7/6/5 mg/kg"]) < 1.1)
```

## Exposure-response curves

```{r fig-er, fig.cap = "Exposure-response for the two statistically significant endpoints (replicates Figure 4 of Chen 2025). Vertical lines mark the mean AUC(0-30min) achieved under each regimen in Table S2."}
auc_grid <- seq(5, 25, by = 0.25)
er_curves <- er_published |>
  dplyr::filter(model %in% c("Chen_2025_hemoporfin_investigator_rating",
                             "Chen_2025_hemoporfin_patient_rating")) |>
  dplyr::rowwise() |>
  dplyr::reframe(
    endpoint = endpoint,
    auc = auc_grid,
    prob = vapply(auc_grid, \(a) solve_er(er_mods[[model]], output, a), 0)
  )

ggplot(er_curves, aes(auc, 100 * prob, colour = endpoint)) +
  geom_vline(xintercept = table_s2$auc, linetype = 3, colour = "grey50") +
  geom_line(linewidth = 0.9) +
  labs(x = "AUC(0-30min) (h*ug/mL)", y = "Probability of a high rating (%)",
       colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

The vertical lines show why the paper argues the exposure gap matters clinically:
moving from the empirical 5 mg/kg regimen to any of the three proposed regimens
crosses the steep part of both curves, roughly doubling the predicted probability of
a high efficacy rating.

# Variability with between-subject random effects

The checks above deliberately use typical values so that they cannot depend on
rxode2's random draw. This section adds the published IIV to compare the *spread* of
exposures, which is the quantity Chen 2025 reports as a %CV.

```{r iiv-cohort}
sim_ped_iiv <- solve_cohort(pediatric, 5, model = pk_mod)
sim_ad_iiv  <- solve_cohort(adult, 5, model = pk_mod)

iiv_cv <- dplyr::bind_rows(sim_ped_iiv, sim_ad_iiv) |>
  dplyr::group_by(cohort, id) |>
  dplyr::summarise(cmax = max(Cc_ug),
                   auc = sum(diff(time) * (head(Cc_ug, -1) + tail(Cc_ug, -1)) / 2),
                   .groups = "drop") |>
  dplyr::group_by(cohort) |>
  dplyr::summarise(`Cmax CV%` = 100 * sd(cmax) / mean(cmax),
                   `AUC0-30min CV%` = 100 * sd(auc) / mean(auc), .groups = "drop")

iiv_cv |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), \(x) round(x, 1))) |>
  dplyr::rename(Cohort = cohort) |>
  knitr::kable(caption = "Simulated variability with IIV (Chen 2025 Results 3.4 reports Cmax 24.8% pediatric / 20.4% adult and AUC 13.5% / 11.9%).")

# A cohort-derived spread depends on the random draw and therefore on the solver
# thread count, so this is bounded by magnitude only -- never by an ordering or
# a sign. The published values span 11.9-24.8%; 45% still goes red if an IIV
# variance were transcribed as an SD (which would roughly triple these).
stopifnot(all(iiv_cv$`Cmax CV%` < 45), all(iiv_cv$`AUC0-30min CV%` < 45))
```

The simulated spreads land between the two published columns, and their *pattern*
is again diagnostic of the observed-versus-EBE split discussed above. Chen 2025
reports a Cmax %CV roughly twice its AUC %CV (24.8 against 13.5 in children), whereas
the model -- in which both metrics are driven by the same three random effects over
the same window -- produces near-identical spreads (18.5 against 18.6). An observed
Cmax carries assay and residual variability that an empirical Bayes AUC does not, and
Bayesian shrinkage pulls the AUC estimates toward the typical value; the two published
columns are therefore not on a common footing, exactly as with the mean values.

# Assumptions and deviations

**Values taken from the final table, not the control stream.** Data S2 prints a
`$THETA` / `$OMEGA` block whose numbers (CL 11.2, Q2 0.374, Q3 0.129, V1 3.5,
V2 0.668, V3 1.33, `theta1` 0.442, `sigma_prop` 0.268, `sigma_add` 0.236,
`theta_study` -0.151) are *initial* estimates. The control stream supplies the model
**structure**; all values come from Table 2.

**IIV scale.** Table 2 reports between-subject variability as a %CV. It is encoded as
a log-scale variance with `omega^2 = log(CV^2 + 1)`. The alternative reading
(`omega^2 = CV^2`) differs by under 1.5% in variance at these magnitudes and does not
affect any conclusion here. The Data S2 `$OMEGA` initial values (0.0178, 0.0404,
0.048) confirm the reported column is a CV and the `$OMEGA` entries are variances --
read as SDs they would imply CVs near 2%, an order of magnitude away from Table 2.

**Zero-variance etas are omitted, not fixed at zero.** The control stream carries an
ETA on all six disposition parameters but fixes the Q3, V2 and V3 elements to `0 FIX`,
which is why Table 2 reports IIV for CL, Q2 and V1 only. Those three are omitted from
`ini()` rather than written as `fixed(0)`: a zero diagonal makes OMEGA singular and
rxode2's Cholesky sampler then fails.

**Concentration units.** The control stream sets `S1 = V1/1000`, which with doses in
mg and volumes in L puts predicted concentrations -- and hence the 0.240 ng/mL
additive residual and the 2 ng/mL LLOQ -- on the **ng/mL** scale. The model therefore
computes `Cc <- 1000 * central / vc`. The paper reports exposure metrics in ug/mL and
h\*ug/mL, so this vignette divides by 1000 before computing AUC.

**The M3 likelihood is not reproduced.** A large proportion of the pediatric 24-hour
samples were below the LLOQ, and the source was fitted with Beal's M3 method: the
control stream's `$ERROR` block switches to `F_FLAG=1` and a `PHI()` censoring
likelihood below 2 ng/mL. Only the above-LLOQ branch -- the combined proportional
plus additive residual -- is encoded, because M3 is an *estimation* device with no
counterpart in a forward simulation. Simulating an observation below 2 ng/mL is
outside the assay's range and should be treated as censored by the user.

**The virtual cohort is reconstructed, not published.** Chen 2025 gives medians and
ranges (Table 1), not per-subject covariates. The deterministic split-lognormal
lattice reproduces each arm's median, minimum and maximum exactly and gives weight,
height and age one shared lattice position per subject so that BMI stays coherent, but
the true joint distribution of the 24 patients is unknown. Sex is assigned in an
alternating 50/50 pattern rather than the actual 10F/14M and 8F/8M splits.

**Fat-free mass equation.** Chen 2025 cites reference [20] for FFM, which is
Al-Sallami et al. 2015 -- the *pediatric* equation -- not the adult Janmahasatian
equation. Check 1 confirms this choice: applying Al-Sallami to the published adult
demographics returns a median FFM within 1.5 kg of the 46.1 kg the paper fixed.

**Body surface area formula.** Table 3's 190 mg/m^2 regimen requires a BSA formula
that Chen 2025 does not state. Du Bois is used here. Mosteller changes the mean
pediatric BSA in this cohort by well under a percent and does not change any
conclusion.

**The study effect is applied to the pediatric arm.** `STUDY_PEDIATRIC = 1` gives
`F1 = 0.825`. Chen 2025 states that its dose-selection simulations proceeded *without*
correcting for this effect, i.e. the offset was left in, which is what is done here.
A user wishing to see the pure allometric prediction should set the covariate to 0.

**Cmax is a documented deviation, and the disagreement is internal to the paper.**
Chen 2025 reports the pediatric-versus-adult gap as 18.7% for Cmax and 30.5% for
AUC(0-30min). In a linear model over a fixed window these two must move together, and
in this simulation they do -- 30.6% and 30.8%, agreeing to 0.2 percentage points.
Table 3 shows the same split: its published Cmax and AUC bias columns differ by about
14 percentage points for the flat 6 mg/kg regimen, whereas the simulated columns agree
to 0.1. Figure 2's caption resolves it -- panel (a) compares **observed** Cmax while
panel (b) compares **EBEs** of AUC(0-30min). An observed peak depends on the sampling
schedule and on the actual infusion duration, which the Data S2 example data set shows
varied between subjects (16 minutes for subject 101, 20 for subject 102). The Cmax
rows are therefore reported and excluded from the gate rather than tuned; every AUC
row, and the ranking of all three proposed regimens, reproduces.

**One Table S2 cell appears to be a typographical error.** Seven of the eight
published probabilities in Table S2 are reproduced to better than 0.1 percentage
points by `expit(beta0 + beta1 * AUC)` using the paper's own Table S1 coefficients and
its own tabulated AUC values. The eighth -- the patient rating under the stratified
7/6/5 mg/kg regimen -- is published as 90.0% where the arithmetic gives 90.9%. The
investigator cell in the same column reproduces exactly (88.2%), so the AUC input for
that column is right and the discrepancy is confined to the one number. A transposed
digit (90.0 for 90.9) is the most likely explanation. The value is reported as
published and bounded at 1.1 percentage points rather than corrected.

**The three non-significant endpoints are shipped but should not be extrapolated.**
Only the investigator- and patient-rating endpoints reached significance. The
`almost_cured` model in particular has an intercept of -29.7 and a slope of 1.98 with a
95% CI of 0.372 to 5.63, the signature of near-separation in 24 patients; its fitted
curve is effectively a step function. The `any_improvement` model has the only
negative slope of the five, which is a sampling artefact rather than evidence that
exposure harms response. Both are extracted verbatim, because Table S1 reports point
estimates for all five endpoints and tuning a published estimate toward an expected
direction is not permitted.

**Placeholder residual on the exposure-response models.** The source likelihood is
Bernoulli, which has no sigma, and no random effects were estimated. Each ER model
carries `addSd_prob_* <- fixed(0.001)` purely so rxode2 has an error model to attach
to the typical-value probability. It is not a published quantity.

**Observation-variable naming.** The five ER models use `prob_<endpoint>` as the
observation variable rather than `Cc`. `Cc` is reserved for drug concentrations, and
these models have no PK layer and no ODE -- the output is a probability. This matches
the existing `Chen_2021_lorlatinib_*` family in nlmixr2lib. `prob_<endpoint>` is itself
a canonical observation form (operator ruling 2026-09-11, validated by
`probOutputRegex`), so this is a convention the models follow rather than a deviation
they take, and `checkModelConventions()` reports no warning for it.

**No covariates other than size and study were tested.** Chen 2025 states explicitly
(Discussion, limitations) that with 24 pediatric patients no covariate screening
beyond allometry and the study effect was performed. Nothing is inferred here about
sex, race, renal or hepatic function.
