---
title: "Avadomide (Cheng 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Avadomide (Cheng 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Cheng Y, Chen J, Pourdehnad M, Zhou S, Li Y. Population
  Pharmacokinetics of CC-122. Clin Pharmacol. 2021;13:61-71.
  doi:10.2147/CPAA.S310604.
* Article: <https://doi.org/10.2147/CPAA.S310604> (open access)

Avadomide (CC-122) is an oral cereblon-modulating agent. The packaged model
`Cheng_2021_avadomide` is the final population PK model of Cheng 2021: a
two-compartment model with first-order absorption after an absorption lag
time (ALAG = 0.246 h) and first-order elimination. The peripheral volume was
fixed at 10 L. Apparent clearance depends on creatinine clearance (linear,
centred at 94.42 mL/min) and on four tumor types (DLBCL, PCNSL, other solid
tumor and multiple myeloma); the apparent central volume depends on body
weight (linear, centred at 74.5 kg), female sex and seven tumor types. Healthy
male subjects are the reference. Residual error is additive on log-transformed
concentrations, encoded as `lnorm()`.

# Population

Cheng 2021 pooled 298 subjects from three studies (Table 1):

* **CC-122-CP-002 Part 1** (n = 30): single ascending oral doses in healthy
  adults.
* **CC-122-CP-005** (n = 48): single oral dose in subjects with mild, moderate
  or severe renal impairment and matched healthy subjects.
* **CC-122-ST-001** (NCT01421524, n = 220): phase 1a/b dose finding in
  patients with advanced solid tumors, non-Hodgkin lymphoma (NHL) or multiple
  myeloma (MM).

Median (range) age was 59.5 (20-91) years, body weight 74.5 (39.8-159.0) kg
and creatinine clearance 94.4 (9.0-321.2) mL/min; 37.9% were female. Tumor
types were healthy (no tumor) 78 (26.2%), DLBCL 60 (20.1%), GBM 44 (14.8%),
NHL 30 (10.1%), MM 29 (9.7%), HCC 27 (9.1%), other solid tumor 19 (6.4%),
brain cancer 6 (2.0%) and primary CNS lymphoma (PCNSL) 5 (1.7%). Doses ranged
from 0.5 to 15 mg. Race and region were not reported.

The same information is available programmatically via
`readModelDb("Cheng_2021_avadomide")()$population`.

# Source trace

Every `ini()` value carries an in-file comment in
`inst/modeldb/specificDrugs/Cheng_2021_avadomide.R`. The table below collects
them.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` (Ka) | log(4.14) 1/h | Table 2, TVKa |
| `ltlag` (ALAG) | log(0.246) h | Table 2, TVALAG |
| `lcl` (CL/F) | log(3.63) L/h | Table 2, TVCL/F |
| `lvc` (V2/F) | log(36.2) L | Table 2, TVV2/F |
| `lq` (Q/F) | log(1.38) L/h | Table 2, TVQ/F |
| `lvp` (V3/F) | fixed(log(10)) L | Table 2, TVV3/F '10 Fix'; Results, Structural model |
| `e_crcl_cl` | 0.007 per mL/min | Table 2 and footnote b |
| `e_tumtp_dlbcl_cl`, `e_tumtp_pcnsl_cl`, `e_tumtp_other_cl`, `e_tumtp_myelo_cl` | -0.647, -0.692, -0.364, 0.309 | Table 2 and footnote b |
| `e_wt_vc` | 0.009 per kg | Table 2 and footnote c |
| `e_sexf_vc` | -0.179 | Table 2 and footnote c ('(1 - 0.179) if female') |
| `e_tumtp_*_vc` (DLBCL, PCNSL, NHL, HCC, GBM, MM, brain cancer) | 0.476, 0.846, 0.344, 0.521, 0.480, 0.682, 0.415 | Table 2 and footnote c |
| `etalcl`, `etalvc`, `etalka`, `etalq` | 0.2725, 0.0681, 2.0506, 0.9624 | Table 2 CV% squared (52.2%, 26.1%, 143.2%, 98.1%) |
| `expSd` | 0.3033 | Table 2, sigma^2 (log additive) = 0.092 |
| CL/F equation | `3.63 * (1 + 0.007 * (CLcr - 94.42)) * prod(1 + theta_k * Z_k)` | Table 2 footnote b; Methods Eqs. 1 and 3 |
| V2/F equation | `36.2 * (1 + 0.009 * (BW - 74.5)) * (1 - 0.179)^female * prod(1 + theta_k * Z_k)` | Table 2 footnote c |
| Structure: `d/dt(depot)`, `d/dt(central)`, `d/dt(peripheral1)`, `alag(depot)` | n/a | Figure 3; Results, Structural model |
| Residual error `Cc ~ lnorm(expSd)` | n/a | Methods: additive error after log-transforming predictions |

# Typical-value covariate effects (Figure 4)

Figure 4 is a forest plot of the covariate effects on CL/F and V2/F. The
Results text reports the numbers behind it, which the model's covariate
factors can be set against directly.

```{r covariate-factors}
mod <- readModelDb("Cheng_2021_avadomide")
theta <- rxode2::rxode(mod)$theta

cl_factor <- tibble::tribble(
  ~category,           ~model,                               ~paper_pct,
  "DLBCL",             1 + theta[["e_tumtp_dlbcl_cl"]],      34.8,
  "PCNSL",             1 + theta[["e_tumtp_pcnsl_cl"]],      30.3,
  "Other solid tumor", 1 + theta[["e_tumtp_other_cl"]],      73.7,
  "MM",                1 + theta[["e_tumtp_myelo_cl"]],      120.0,
  "NHL",               1,                                    89.3,
  "HCC",               1,                                    109.2,
  "GBM",               1,                                    107.9,
  "Brain cancer",      1,                                    131.6
) |>
  dplyr::mutate(model_pct = 100 * model)

cl_factor |>
  dplyr::select(category, model_pct, paper_pct) |>
  dplyr::rename(
    "Tumor type" = category,
    "Model CL/F, % of healthy" = model_pct,
    "Results text, % of healthy" = paper_pct
  ) |>
  knitr::kable(digits = 1, caption = "Typical CL/F by tumor type relative to healthy subjects.")

# Structural gate: the two large, precisely estimated CL/F effects must
# reproduce the Results text to within 2 percentage points. A sign or
# transcription error in either coefficient moves the factor by tens of points.
stopifnot(
  abs(cl_factor$model_pct[cl_factor$category == "DLBCL"] - 34.8) < 2,
  abs(cl_factor$model_pct[cl_factor$category == "PCNSL"] - 30.3) < 2
)
```

The DLBCL and PCNSL factors agree with the text. The remaining percentages in
the text do not come from the Table 2 coefficients: for NHL, HCC, GBM and brain
cancer the text reports 89.3-131.6% of healthy CL/F and then states that these
effects were fixed to 0 in the model, so the text values describe
individual estimates rather than the model. The same holds for other solid
tumor (73.7% in the text, 63.6% from the coefficient) and MM (120.0% versus
130.9%). Figure 4 plots medians of individual estimates, so it is a
descriptive summary and not a readout of Table 2.

```{r covariate-factors-v2}
vc_factor <- tibble::tribble(
  ~category,      ~model,
  "DLBCL",        1 + theta[["e_tumtp_dlbcl_vc"]],
  "PCNSL",        1 + theta[["e_tumtp_pcnsl_vc"]],
  "NHL",          1 + theta[["e_tumtp_nhl_vc"]],
  "HCC",          1 + theta[["e_tumtp_hcc_vc"]],
  "GBM",          1 + theta[["e_tumtp_glio_vc"]],
  "MM",           1 + theta[["e_tumtp_myelo_vc"]],
  "Brain cancer", 1 + theta[["e_tumtp_brain_vc"]],
  "Female",       1 + theta[["e_sexf_vc"]]
)
vc_factor |>
  dplyr::mutate(model = 100 * model) |>
  dplyr::rename("Category" = category, "Model V2/F, % of reference" = model) |>
  knitr::kable(digits = 1, caption = "Typical V2/F relative to healthy male subjects.")
```

The text reports V2/F 3.3% to 57.8% higher in cancer patients and 26.8% higher
in males. The coefficients give 34.4% to 84.6% higher V2/F by tumor type and
1 / (1 - 0.179) = 21.8% higher in males. These are again individual-estimate
summaries (Figure 4B) and are not expected to match the coefficients.

# Virtual cohort

The observed data are not public. The cohorts below approximate the
demographics of each study in Table 1. Body weight and creatinine clearance
are drawn log-normally around each study's median and truncated to its range.
ST-001 tumor types are drawn with the Table 1 counts (60 DLBCL, 5 PCNSL,
19 other solid tumor, 30 NHL, 27 HCC, 44 GBM, 29 MM, 6 brain cancer).
Every subject gets a single 3 mg oral dose: the dose Li 2020 used in the renal
impairment study (CP-005) and within the 0.5-15 mg range of the pooled data.

```{r cohort}
set.seed(2021)

rtrunc_lnorm <- function(n, median, sdlog, lo, hi) {
  pmin(pmax(median * exp(rnorm(n, 0, sdlog)), lo), hi)
}

tumor_levels <- c(
  "DLBCL", "PCNSL", "Other solid tumor", "NHL", "HCC", "GBM", "MM", "Brain cancer"
)
tumor_counts <- c(60, 5, 19, 30, 27, 44, 29, 6)

make_cohort <- function(n, study, wt_med, wt_lo, wt_hi, crcl_med, crcl_sd,
                        crcl_lo, crcl_hi, pct_female, tumor, id_offset) {
  tibble::tibble(
    id = id_offset + seq_len(n),
    study = study,
    WT = rtrunc_lnorm(n, wt_med, 0.2, wt_lo, wt_hi),
    CRCL = rtrunc_lnorm(n, crcl_med, crcl_sd, crcl_lo, crcl_hi),
    SEXF = rbinom(n, 1, pct_female / 100),
    tumor = tumor
  )
}

subjects <- dplyr::bind_rows(
  make_cohort(100, "CP-002", 75.8, 52.9, 95.0, 107.0, 0.2, 76.3, 171.0, 16.7,
              "Healthy", id_offset = 0L),
  make_cohort(100, "CP-005", 79.1, 57.7, 126.0, 90.6, 0.6, 9.0, 152.0, 43.7,
              "Healthy", id_offset = 100L),
  make_cohort(200, "ST-001", 73.0, 39.8, 159.0, 92.1, 0.3, 35.3, 321.2, 39.5,
              sample(rep(tumor_levels, tumor_counts), 200), id_offset = 200L)
) |>
  dplyr::mutate(
    TUMTP_DLBCL = as.integer(tumor == "DLBCL"),
    TUMTP_PCNSL = as.integer(tumor == "PCNSL"),
    TUMTP_OTHER = as.integer(tumor == "Other solid tumor"),
    TUMTP_NHL = as.integer(tumor == "NHL"),
    TUMTP_HCC = as.integer(tumor == "HCC"),
    TUMTP_GLIO = as.integer(tumor == "GBM"),
    TUMTP_MYELO = as.integer(tumor == "MM"),
    TUMTP_BRAIN = as.integer(tumor == "Brain cancer")
  )
stopifnot(!anyDuplicated(subjects$id))

obs_times <- c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 12, 24, 36, 48, 72)
dose_amt <- 3

events <- dplyr::bind_rows(
  subjects |> dplyr::mutate(time = 0, evid = 1L, amt = dose_amt, cmt = "depot"),
  subjects |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(evid = 0L, amt = 0, cmt = "central")
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))
```

# Simulation

```{r simulate}
keep_cols <- c("study", "tumor", "CRCL", "WT", "SEXF")
sim <- rxode2::rxSolve(mod, events = events, keep = keep_cols) |>
  as.data.frame()

mod_typical <- rxode2::zeroRe(mod)
sim_typical <- rxode2::rxSolve(mod_typical, events = events, keep = keep_cols) |>
  as.data.frame()
```

# Replicate published figures

## Figure 2: dose-normalized profiles by study

```{r figure-2, fig.width = 7, fig.height = 4.5}
sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(cn = Cc / dose_amt) |>
  dplyr::group_by(study, time) |>
  dplyr::summarise(
    Q05 = quantile(cn, 0.05), Q50 = median(cn), Q95 = quantile(cn, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50, colour = study, fill = study)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  labs(
    x = "Time after dose (h)", y = "Concentration / dose (ng/mL/mg)",
    colour = "Study", fill = "Study",
    title = "Simulated dose-normalized CC-122 profiles by study (median, 5th-95th)",
    caption = "Compare with Figure 2 of Cheng 2021 (individual observed profiles)."
  )
```

Figure 2 of the paper shows most dose-normalized peaks between about 20 and
60 ng/mL/mg, with the widest spread in ST-001. In the simulation the typical
healthy peak is about 23 ng/mL/mg (Cmax / 3 mg in the NCA table below). ST-001
patients have lower median peaks because most tumor types raise V2/F by 34-85%,
and they decline more slowly because DLBCL and PCNSL cut CL/F by about
two-thirds. CP-005 declines more slowly than CP-002 because of its
renally impaired subjects. The figure shows individual observed profiles, so
the comparison is qualitative.

## Figure 5: VPC of log concentration

```{r figure-5, fig.width = 7, fig.height = 4.5}
# Add the log-additive residual error explicitly (sigma^2 = 0.092) so the
# percentiles are those of simulated observations, as in a VPC.
exp_sd <- theta[["expSd"]]
sim |>
  dplyr::filter(!is.na(Cc), time > 0) |>
  dplyr::mutate(log_obs = log(Cc) + rnorm(dplyr::n(), 0, exp_sd)) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    Q05 = quantile(log_obs, 0.05), Q50 = median(log_obs),
    Q95 = quantile(log_obs, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "firebrick", linewidth = 0.8) +
  labs(
    x = "Time after dose (hours)", y = "Log(CC-122) (ng/mL)",
    title = "Simulated 5th, 50th and 95th percentiles, pooled cohort, 3 mg",
    caption = "Compare with Figure 5 of Cheng 2021 (doses pooled, 0.5-15 mg)."
  )
```

Figure 5 pools every dose level in the dataset, so the simulated 3 mg
percentiles are a shape comparison only. Read by eye, the observed median in
Figure 5 falls by about 1.7 log units between 24 and 48 h, which corresponds to a half-life of
about 10 h. The typical healthy subject in the model has a terminal half-life
of the same size (computed below).

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, study)
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, study) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, study, time, .keep_all = TRUE) |>
  dplyr::arrange(id, study, time)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, study)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | study + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | study + id)
intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_summary <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  dplyr::group_by(study, PPTESTCD) |>
  dplyr::summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median)

nca_summary |>
  dplyr::rename(
    "Study" = study,
    "Cmax (ng/mL)" = cmax,
    "Tmax (h)" = tmax,
    "AUC0-inf (ng*h/mL)" = aucinf.obs,
    "t1/2 (h)" = half.life
  ) |>
  knitr::kable(digits = 2, caption = "Median simulated NCA after a single 3 mg dose.")
```

## Comparison against published NCA

The Introduction of Cheng 2021 summarises the single-dose NCA in healthy adults
(3 to 15 mg): Tmax approximately 1 h and terminal half-life 7.6-8.9 h. The
healthy single-ascending-dose cohort (CP-002) is the matching simulated group.

```{r nca-compare}
published <- tibble::tibble(study = "CP-002", tmax = 1)
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "study",
  params = "tmax",
  units = c(tmax = "h"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Simulated vs. published Tmax (CP-002, healthy). * differs by >20%.")

hl_cp002 <- nca_summary$half.life[nca_summary$study == "CP-002"]
tmax_cp002 <- nca_summary$tmax[nca_summary$study == "CP-002"]

# Typical healthy reference subject: terminal (beta) half-life from the
# micro-constants (CLcr 94.42 mL/min, 74.5 kg, male).
k10 <- exp(theta[["lcl"]]) / exp(theta[["lvc"]])
k12 <- exp(theta[["lq"]]) / exp(theta[["lvc"]])
k21 <- exp(theta[["lq"]]) / exp(theta[["lvp"]])
beta <- ((k10 + k12 + k21) - sqrt((k10 + k12 + k21)^2 - 4 * k10 * k21)) / 2
hl_typical <- log(2) / beta

tibble::tibble(
  Quantity = c(
    "Median simulated t1/2, CP-002 (h)",
    "Typical-subject terminal t1/2 (h)",
    "Published t1/2 range, healthy (h)"
  ),
  Value = c(
    sprintf("%.1f", hl_cp002), sprintf("%.1f", hl_typical), "7.6-8.9"
  )
) |>
  knitr::kable(caption = "Terminal half-life.")

# Gates on medians of 100 simulated subjects. Tmax: lag 0.246 h plus
# Ka = 4.14 1/h puts the typical peak near 1 h; a mis-transcribed lag or Ka
# moves it by more than 0.5 h. Half-life: the typical value is ~10 h, and a
# broken Q/F or V3/F moves it by a factor of 2 or more.
stopifnot(
  abs(tmax_cp002 - 1) < 0.5,
  hl_typical > 7, hl_typical < 13,
  hl_cp002 > 6, hl_cp002 < 14
)
```

Simulated Tmax matches the published value. The model's terminal half-life is
longer than the published 7.6-8.9 h range. That range comes from a
dense-sampling NCA of healthy adults in a separate report cited by the paper;
the popPK model fixed V3/F at 10 L and estimated Q/F with a 49.9% RSE and a
98.1% CV, so its terminal phase is not tightly identified, and the difference
is recorded here rather than adjusted.

## Closed-form AUC check

For a linear model the typical-value AUC0-inf after an oral dose is
Dose / (CL/F). This check runs on the zero-random-effect solve.

```{r auc-closed-form}
cl_typ <- sim_typical |>
  dplyr::distinct(id, .keep_all = TRUE) |>
  dplyr::select(id, study, cl)

conc_typ <- sim_typical |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, study)
conc_typ <- dplyr::bind_rows(
  conc_typ,
  conc_typ |> dplyr::distinct(id, study) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, study, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

nca_typ <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_typ, Cc ~ time | study + id),
  PKNCA::PKNCAdose(dose_df, amt ~ time | study + id),
  intervals = data.frame(start = 0, end = Inf, aucinf.obs = TRUE)
))

auc_chk <- as.data.frame(nca_typ$result) |>
  dplyr::filter(PPTESTCD == "aucinf.obs") |>
  dplyr::select(id, study, auc = PPORRES) |>
  dplyr::left_join(cl_typ, by = c("id", "study")) |>
  dplyr::mutate(
    auc_closed = dose_amt * 1000 / cl,
    pct_diff = 100 * (auc / auc_closed - 1)
  )

auc_chk |>
  dplyr::group_by(study) |>
  dplyr::summarise(
    median_pct_diff = median(pct_diff),
    max_abs_pct_diff = max(abs(pct_diff)),
    .groups = "drop"
  ) |>
  dplyr::rename(
    "Study" = study,
    "Median % difference" = median_pct_diff,
    "Max |% difference|" = max_abs_pct_diff
  ) |>
  knitr::kable(digits = 2, caption = "PKNCA AUC0-inf vs Dose / (CL/F), typical values.")

# Trapezoidal AUC over a 72 h grid plus log-linear extrapolation differs from
# the exact integral by a few percent; a wrong CL/F, dose or unit moves it by
# tens of percent.
stopifnot(
  abs(median(auc_chk$pct_diff)) < 5,
  quantile(abs(auc_chk$pct_diff), 0.9) < 10
)
```

# Renal function

Cheng 2021 identified creatinine clearance as the covariate supporting dose
adjustment in renal impairment. Because the effect is linear, typical CL/F
falls to 40% of the reference at the lowest observed CLcr (9 mL/min).

```{r renal, fig.width = 6, fig.height = 4}
crcl_grid <- tibble::tibble(CRCL = seq(9, 180, by = 1)) |>
  dplyr::mutate(
    cl = exp(theta[["lcl"]]) * (1 + theta[["e_crcl_cl"]] * (CRCL - 94.42))
  )
ggplot(crcl_grid, aes(CRCL, cl)) +
  geom_line() +
  geom_vline(xintercept = c(30, 60, 90), linetype = "dashed", colour = "grey50") +
  labs(
    x = "Creatinine clearance (mL/min)", y = "Typical CL/F, healthy (L/h)",
    title = "Typical CL/F vs creatinine clearance"
  )
```

# Assumptions and deviations

- **IIV scale.** Table 2 reports IIV as CV%. The bootstrap-median column holds
  the variances, and for V2/F, Ka and Q the printed CV% equals
  sqrt(omega^2) exactly (sqrt(0.068) = 26.1%, sqrt(2.05) = 143.2%,
  sqrt(0.962) = 98.1%). The variances are therefore taken as (CV%/100)^2. For
  CL/F this gives 0.2725, against a bootstrap median of 0.261; the bootstrap
  interval (0.168-0.308) contains the point estimate. No IIV covariances were
  reported, so OMEGA is diagonal.
- **IIV on Q.** Table 2 reports IIV on "Q" with no /F; it is applied to the
  apparent intercompartmental clearance Q/F, the only Q in the model.
- **Effects held at zero.** The Results state that tumor-type effects of NHL,
  HCC, GBM and brain cancer on CL/F, and of other solid tumor on V2/F, were
  fixed to 0 in the final model. They are omitted from the model, which is
  numerically the same as encoding them as zero.
- **MM on V2/F.** Table 2 prints 0.682 as the estimate and 0.662 as the
  bootstrap median. The model uses the estimate, which also appears in the
  footnote c equation.
- **Creatinine clearance.** The estimating equation (for example
  Cockcroft-Gault) and any BSA normalization are not stated. Values are used
  in mL/min as in Table 1. The Results text writes the renal-function bands as
  "mL/hr", which is a typo for mL/min.
- **Tumor-type coding.** The paper's tumor-type levels are mutually
  exclusive. DLBCL and PCNSL have their own indicators, so `TUMTP_NHL` marks
  only the other NHL patients. `TUMTP_BRAIN` marks brain cancer other than GBM
  (`TUMTP_GLIO`) and PCNSL. `TUMTP_OTHER` is the paper's "other solid tumor".
  Healthy subjects, including the renal-impairment subjects of CP-005, have
  every indicator at 0.
- **Enantiomers.** CC-122 is racemic. CP-002 and CP-005 used an achiral assay
  (total CC-122); ST-001 used a chiral assay for the R and S enantiomers. The
  paper does not say how the enantiomer data entered the model; the model
  describes total CC-122.
- **Virtual cohort.** Body-weight and creatinine-clearance spreads (log-normal
  SD 0.2-0.6, truncated to the Table 1 ranges) and the uniform 3 mg single dose
  are choices made for this vignette; the paper does not give per-subject
  doses or regimens.
- **Terminal half-life.** The typical-subject terminal half-life (about 10 h)
  is longer than the 7.6-8.9 h quoted from a separate healthy-subject NCA. The
  difference is reported above and the parameters were not adjusted.
- No correction notice for this article was found in Europe PMC as of
  2026-09-28.
