---
title: "Luspatercept in beta-thalassemia (Chen 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Luspatercept in beta-thalassemia (Chen 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Chen_2021_luspatercept"))$reference`
* Article: [J Clin Pharmacol. 2021;61(1):52-63](https://doi.org/10.1002/jcph.1696) (open access, PMC7754485)

Chen 2021 reports a population PK model of subcutaneous luspatercept in
adults with beta-thalassemia, followed by exposure-response analyses that use
the model's individual apparent clearance to derive each patient's exposure.
The paper contributes five models to nlmixr2lib:

| Model | Endpoint | Source |
|---|---|---|
| `Chen_2021_luspatercept` | Serum luspatercept concentration (one-compartment PK) | Table 2; Results equations |
| `Chen_2021_luspatercept_hb_change` | Average hemoglobin change from baseline, weeks 1-3 (linear) | Figure 2A |
| `Chen_2021_luspatercept_hb_response` | Probability of a >= 1 g/dL day-21 hemoglobin increase (logistic) | Figure 2B |
| `Chen_2021_luspatercept_teae_grade3` | Probability of a grade >= 3 TEAE in cycles 1-2 (logistic) | Figure 4A |
| `Chen_2021_luspatercept_bone_pain` | Probability of bone pain of any grade in cycles 1-2 (logistic) | Figure 4B |

The four exposure-response models have no PK layer: each takes the
starting-dose steady-state AUC over the 21-day dosing interval as the
covariate column `AUC_LUSP` (ug*day/mL). Chen 2021 computed that AUC as
starting dose / individual CL/F, which is what this vignette does with the
PK model.

# Population

The PK population comprised 285 adults with beta-thalassemia: 64 from the
phase 2 dose-finding/expansion study A536-04 (NCT01749540; continued in the
extension A536-06) and 221 from the phase 3 BELIEVE study ACE-536-B-THAL-001
(NCT02604433). From Chen 2021 Table 1: median age 32 years (18-66), median
weight 57.1 kg (34.1-97.0), 56.8% female, 63.5% White and 28.8% Asian,
median baseline albumin 46.0 g/L (30.0-56.0), median baseline RBC
transfusion burden 14.1 units/24 weeks (0-34.0), 59.6% splenectomised, and
86.0% with normal renal function. Luspatercept was given subcutaneously once
every 3 weeks at 0.2-1.25 mg/kg; 79.8% of patients started at 1 mg/kg and 34%
escalated to 1.25 mg/kg during the first year. The analysis used 3680
quantifiable serum concentrations collected 5-610 days after the first dose.

# Source trace

| Quantity | Value | Source |
|---|---|---|
| Structure: 1-compartment, first-order absorption and elimination | -- | Results, 'Luspatercept Population PK Model' |
| CL/F | 0.532 L/day | Table 2 |
| V1/F | 8.39 L | Table 2 |
| Ka | 0.409 1/day | Table 2 |
| CL/F covariate equation | 0.532 x (WT/70)^0.806 x (ALB/46)^-0.881 x exp(-0.0118 x [RBCT - 14]) | Results equation |
| V1/F covariate equation | 8.39 x (WT/70)^0.705 x exp(-0.0141 x [RBCT - 14]) | Results equation |
| IIV CL/F, V1/F | 34.7%, 27.6% (exponential) | Table 2; Methods |
| Residual error | 20.8% (additive on log-transformed data) | Table 2; Methods |
| Hb change: slope | 0.011 g/dL per ug*day/mL (R = 0.72) | Figure 2A panel annotation |
| Hb change: intercept | -0.327 g/dL | Digitized from the Figure 2A fitted line |
| Hb change: residual SD | 0.511 g/dL | OLS refit of the digitized Figure 2A points |
| Hb >= 1 g/dL: slope | log(3.73)/50 = 0.02633 per ug*day/mL | Figure 2B OR per 50 units of AUCss |
| Hb >= 1 g/dL: intercept | -3.593 | Digitized from the Figure 2B fitted curve |
| TEAE >= grade 3, cycles 1-2: intercept, slope | -2.360, -0.0001916 per ug*day/mL | Digitized from the Figure 4A fitted line |
| Bone pain, cycles 1-2: intercept, slope | -1.144, -0.002191 per ug*day/mL | Digitized from the Figure 4B fitted line |

## How the figure-derived coefficients were obtained

The figures in the published PDF are vector graphics, so the fitted lines,
the scatter points and the axis tick marks are stored as exact coordinates
rather than pixels. The maintainers extracted those coordinates, mapped each
axis by linear regression on its tick marks, and read the coefficients from
the mapped lines. No curve was fitted by eye. The chunk below repeats the
arithmetic from the extracted coordinates (PDF points, y increasing downward)
so that it can be audited.

```{r digitization}
axis_map <- function(tick_pos, tick_val) {
  fit <- stats::lm(tick_val ~ tick_pos)
  function(pos) unname(stats::coef(fit)[1] + stats::coef(fit)[2] * pos)
}

# Figure 2A (page 7): x ticks 0-200, y ticks -2 to 3.
x2a <- axis_map(c(80.74, 124.64, 168.845, 212.755, 256.225), seq(0, 200, 50))
y2a <- axis_map(c(187.475, 166.345, 145.415, 124.565, 103.38, 82.375), -2:3)
line2a <- data.frame(auc = x2a(c(94.13, 267.33)), dhb = y2a(c(147.925, 104.115)))
slope2a_drawn <- diff(line2a$dhb) / diff(line2a$auc)
mid2a <- colMeans(line2a)
int2a <- unname(mid2a["dhb"] - 0.011 * mid2a["auc"])

# Figure 2B (page 7): x ticks 0-200, y ticks 0-1. Three on-curve Bezier nodes.
x2b <- axis_map(c(328.155, 372.06, 416.27, 460.175, 503.645), seq(0, 200, 50))
y2b <- axis_map(c(187.475, 166.345, 145.415, 124.565, 103.38, 82.375), seq(0, 1, 0.2))
node2b <- data.frame(auc = x2b(c(341.325, 434.83, 514.275)), p = y2b(c(183.315, 145.235, 95.515)))
slope2b <- log(3.73) / 50
int2b <- stats::qlogis(node2b$p[2]) - slope2b * node2b$auc[2]
free2b <- stats::coef(stats::lm(stats::qlogis(p) ~ auc, node2b))

# Figures 4A and 4B (page 9): x ticks 0-300, y ticks 0-1; two line end nodes.
x4 <- axis_map(c(89.67, 111.49, 133.34, 155.285, 176.96, 199.12, 220.52), seq(0, 300, 50))
two_point_logit <- function(xp, yp, ytick) {
  y4 <- axis_map(ytick, seq(0, 1, 0.2))
  auc <- x4(xp)
  lg <- stats::qlogis(y4(yp))
  slope <- diff(lg) / diff(auc)
  c(intercept = lg[1] - slope * auc[1], slope = slope)
}
c4a <- two_point_logit(c(95.49, 221.05), c(185.045, 185.505),
                       c(194.22, 172.75, 151.095, 129.2, 107.81, 85.57))
c4b <- two_point_logit(c(95.49, 221.055), c(346.46, 356.75),
                       c(371.935, 350.465, 328.81, 306.915, 285.52, 263.28))

digit <- tibble::tribble(
  ~Coefficient,                          ~Derived,        ~`In model`,
  "Fig 2A slope of drawn line",          slope2a_drawn,   0.011,
  "Fig 2A intercept (printed slope)",    int2a,           -0.327,
  "Fig 2B intercept (printed OR slope)", int2b,           -3.593,
  "Fig 2B free-fit slope (3 nodes)",     free2b[[2]],     0.02633,
  "Fig 4A intercept",                    c4a[["intercept"]], -2.360,
  "Fig 4A slope",                        c4a[["slope"]],  -0.0001916,
  "Fig 4B intercept",                    c4b[["intercept"]], -1.144,
  "Fig 4B slope",                        c4b[["slope"]],  -0.002191
)
knitr::kable(digit, digits = 5)

stopifnot(
  # The printed values that the drawn lines must agree with.
  abs(slope2a_drawn - 0.011) < 0.0006,        # printed to 2 significant figures
  abs(free2b[[2]] / slope2b - 1) < 0.02,      # printed OR 3.73 per 50 units
  # The packaged coefficients are the derived ones, rounded.
  abs(int2a - (-0.327)) < 5e-4, abs(int2b - (-3.593)) < 5e-4,
  abs(c4a[["intercept"]] - (-2.360)) < 5e-3, abs(c4a[["slope"]] / -0.0001916 - 1) < 1e-3,
  abs(c4b[["intercept"]] - (-1.144)) < 5e-4, abs(c4b[["slope"]] / -0.002191 - 1) < 1e-3
)
```

Two independent confirmations make the digitization trustworthy. In
Figure 2B, fitting both logistic coefficients freely to the curve's three
nodes gives a slope within 1% of the value implied by the printed odds
ratio. In Figure 2A, the 33 visible scatter points (34 patients; one marker
is hidden behind another) reproduce the printed slope and correlation:

```{r figure-2a-points}
fig2a <- tibble::tribble(
  ~auc, ~dhb, ~dose,
  15.2, -0.33, "0.2", 15.7, -0.40, "0.2", 16.9, 0.00, "0.2", 19.3, -0.19, "0.2",
  24.5, 0.72, "0.2", 26.2, 0.41, "0.2", 28.3, -0.36, "0.4", 28.9, 0.22, "0.4",
  33.8, 0.27, "0.4", 39.1, -1.34, "0.4", 39.4, 0.41, "0.6", 52.9, 0.72, "0.6",
  54.9, 0.79, "0.6", 55.3, 0.06, "0.8", 62.5, 0.48, "0.4", 67.0, -0.69, "0.8",
  71.7, 0.22, "0.6", 76.4, 0.22, "0.6", 82.1, 1.40, "0.8", 88.9, 0.65, "0.8",
  88.9, 1.00, "1", 92.2, 1.00, "0.8", 93.1, 0.86, "0.8", 94.4, 0.52, "0.8",
  98.6, 0.69, "0.8", 99.8, 1.47, "0.8", 110.6, 0.95, "0.8", 116.8, 0.83, "0.8",
  122.8, 0.13, "1.25", 135.3, 0.97, "0.8", 163.6, 1.53, "0.8", 176.7, 1.04, "1",
  212.9, 2.51, "0.8"
)
refit <- stats::lm(dhb ~ auc, fig2a)
refit_tab <- tibble::tibble(
  Statistic = c("Slope (g/dL per ug*day/mL)", "Intercept (g/dL)", "Correlation R", "Residual SD (g/dL)"),
  Published = c("0.011", "not printed", "0.72", "not printed"),
  `Refit of digitized points` = c(stats::coef(refit)[[2]], stats::coef(refit)[[1]],
                                  stats::cor(fig2a$auc, fig2a$dhb), summary(refit)$sigma)
)
knitr::kable(refit_tab, digits = 4)
stopifnot(
  abs(stats::coef(refit)[[2]] - 0.011) < 0.0006,
  abs(stats::cor(fig2a$auc, fig2a$dhb) - 0.72) < 0.01
)
```

# Virtual cohort

Individual covariates are not published, so the cohort below is drawn to
match the Table 1 medians and ranges. Weight is log-normal around 57.1 kg
and albumin normal around 46 g/L; baseline transfusion burden is normal
around 14.1 units/24 weeks. Values outside the published ranges are
redrawn rather than clamped, so the ranges act as definitions and do not
pile subjects up at the limits.

```{r cohort}
set.seed(20210101)
rxode2::rxSetSeed(20210101)

draw_in_range <- function(n, draw, lo, hi) {
  out <- draw(n)
  bad <- out < lo | out > hi
  while (any(bad)) {
    out[bad] <- draw(sum(bad))
    bad <- out < lo | out > hi
  }
  out
}

n_per_arm <- 200
arms <- c(1, 1.25)
cohort <- tibble::tibble(
  id = seq_len(n_per_arm * length(arms)),
  dose_mgkg = rep(arms, each = n_per_arm),
  WT = draw_in_range(n_per_arm * length(arms), function(n) exp(rnorm(n, log(57.1), 0.2)), 34.1, 97.0),
  ALB = draw_in_range(n_per_arm * length(arms), function(n) rnorm(n, 46, 4), 30, 56),
  RBCT_BL = draw_in_range(n_per_arm * length(arms), function(n) rnorm(n, 14.1, 6), 0, 34)
) |>
  dplyr::mutate(
    dose_mg = dose_mgkg * WT,
    treatment = paste(dose_mgkg, "mg/kg q3w")
  )

cohort |>
  dplyr::summarise(dplyr::across(c(WT, ALB, RBCT_BL), list(median = median, min = min, max = max))) |>
  tidyr::pivot_longer(dplyr::everything()) |>
  knitr::kable(digits = 1, caption = "Virtual-cohort covariate summary (Table 1: WT 57.1 [34.1-97.0] kg, ALB 46.0 [30.0-56.0] g/L, RBCT 14.1 [0-34.0] units/24 weeks).")
```

# Simulation

Ten once-every-3-week doses (210 days) are simulated. The mean half-life is
about 11 days, so the tenth interval (days 189-210) is at steady state even
for subjects in the slow tail of the clearance distribution.

```{r simulate}
mod <- readModelDb("Chen_2021_luspatercept")
tau <- 21
n_dose <- 10
obs_times <- sort(unique(c(seq(0, n_dose * tau, by = 1), (n_dose - 1) * tau + c(0.25, 0.5))))

events <- dplyr::bind_rows(
  cohort |>
    dplyr::transmute(id, time = 0, amt = dose_mg, evid = 1L, cmt = "depot",
                     ii = tau, addl = n_dose - 1L),
  tidyr::expand_grid(id = cohort$id, time = obs_times) |>
    dplyr::mutate(amt = 0, evid = 0L, cmt = "central", ii = 0, addl = 0L)
) |>
  dplyr::left_join(dplyr::select(cohort, id, WT, ALB, RBCT_BL, treatment), by = "id") |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim <- rxode2::rxSolve(mod, dplyr::select(events, -treatment), returnType = "data.frame") |>
  dplyr::mutate(id = as.integer(as.character(id))) |>
  dplyr::left_join(dplyr::select(cohort, id, treatment), by = "id")
stopifnot(dplyr::n_distinct(sim$id) == nrow(cohort), !anyNA(sim$Cc), !anyNA(sim$treatment))
```

## Figure 1B -- concentration versus time after dose

Chen 2021 Figure 1B is a visual predictive check of concentration against
time after each dose (0-55 days) with a median near 8-10 ug/mL at the 1 mg/kg
steady-state peak. The simulated steady-state profiles below show the same
quantity.

```{r figure-1b}
sim |>
  dplyr::filter(time >= (n_dose - 1) * tau) |>
  dplyr::mutate(tad = time - (n_dose - 1) * tau) |>
  dplyr::group_by(treatment, tad) |>
  dplyr::summarise(
    Q05 = stats::quantile(Cc, 0.05), Q50 = stats::median(Cc), Q95 = stats::quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~treatment) +
  labs(x = "Time after dose (days)", y = "Luspatercept (ug/mL)",
       caption = "Simulated 5th-50th-95th percentiles over the 10th dosing interval; compare Chen 2021 Figure 1B.") +
  theme_minimal()
```

# PKNCA validation

```{r pknca}
ss_start <- (n_dose - 1) * tau
nca_conc <- sim |>
  dplyr::filter(time >= ss_start, !is.na(Cc)) |>
  dplyr::mutate(time_ss = time - ss_start) |>
  dplyr::select(id, time_ss, Cc, treatment)
nca_dose <- cohort |>
  dplyr::transmute(id, time_ss = 0, amt = dose_mg, treatment)

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ time_ss | treatment + id)
dose_obj <- PKNCA::PKNCAdose(nca_dose, amt ~ time_ss | treatment + id)
intervals <- data.frame(start = 0, end = tau, cmax = TRUE, tmax = TRUE, auclast = TRUE, half.life = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_df <- as.data.frame(nca_res$result)
summary(nca_res)
```

## Comparison against published values

Chen 2021 reports a mean elimination half-life of about 11 days, a mean
AUCss of 129 ug*day/mL after 1 mg/kg (among the patients of the
exposure-hemoglobin analysis), and a 36% coefficient of variation for AUCss
(Results).

```{r comparison}
auc_1 <- nca_df |> dplyr::filter(treatment == "1 mg/kg q3w", PPTESTCD == "auclast")
hl_1 <- nca_df |> dplyr::filter(treatment == "1 mg/kg q3w", PPTESTCD == "half.life")
typical_hl <- log(2) * 8.39 / 0.532
cmp <- tibble::tibble(
  `NCA parameter` = c("AUCss (0-21 d), mean", "AUCss, CV%", "t1/2, mean", "t1/2, typical subject (ln2 V/CL)"),
  Published = c("129 ug*day/mL", "36%", "~11 days", "~11 days"),
  Simulated = c(sprintf("%.0f ug*day/mL", mean(auc_1$PPORRES)),
                sprintf("%.0f%%", 100 * stats::sd(auc_1$PPORRES) / mean(auc_1$PPORRES)),
                sprintf("%.1f days", mean(hl_1$PPORRES, na.rm = TRUE)),
                sprintf("%.1f days", typical_hl))
)
knitr::kable(cmp)

stopifnot(
  abs(mean(auc_1$PPORRES) / 129 - 1) < 0.15,
  abs(typical_hl - 11) < 0.5,
  abs(stats::median(hl_1$PPORRES, na.rm = TRUE) / 11 - 1) < 0.2
)
```

The simulated mean AUCss depends on the assumed covariate distributions; the
published 129 ug*day/mL refers to the phase 2 subset of the exposure-hemoglobin
analysis, so agreement within 15% is the appropriate check rather than an
exact match.

## Figure 1C -- clinical relevance of covariates

Figure 1C compares median exposure at covariate values below the 10th or
above the 90th percentile against patients within the 10th-90th percentiles,
under weight-based (1.25 mg/kg) and fixed (71 mg) dosing. Because
luspatercept PK is linear, AUCss = dose / CL/F exactly, so the comparison can
be made deterministically at the percentile thresholds printed on the
Figure 1C axis (albumin 42 and 50 g/L; transfusion burden 7 and 20
units/24 weeks; weight 45 and 74 kg), holding the other covariates at their
medians.

```{r figure-1c}
cl_typ <- function(WT, ALB, RBCT) 0.532 * (WT / 70)^0.806 * (ALB / 46)^-0.881 * exp(-0.0118 * (RBCT - 14))
ref <- list(WT = 57.1, ALB = 46, RBCT = 14.1)
auc_at <- function(WT = ref$WT, ALB = ref$ALB, RBCT = ref$RBCT, fixed = FALSE) {
  dose <- if (fixed) 71 else 1.25 * WT
  dose / cl_typ(WT, ALB, RBCT)
}
fig1c <- tibble::tribble(
  ~Covariate, ~Dosing, ~value, ~auc,
  "Albumin < 42 g/L", "weight-based", 42, auc_at(ALB = 42),
  "Albumin > 50 g/L", "weight-based", 50, auc_at(ALB = 50),
  "RBCT < 7", "weight-based", 7, auc_at(RBCT = 7),
  "RBCT > 20", "weight-based", 20, auc_at(RBCT = 20),
  "Weight < 45 kg", "weight-based", 45, auc_at(WT = 45),
  "Weight > 74 kg", "weight-based", 74, auc_at(WT = 74),
  "Weight < 45 kg", "fixed 71 mg", 45, auc_at(WT = 45, fixed = TRUE),
  "Weight > 74 kg", "fixed 71 mg", 74, auc_at(WT = 74, fixed = TRUE)
) |>
  dplyr::mutate(
    auc_ref = ifelse(Dosing == "fixed 71 mg", auc_at(fixed = TRUE), auc_at()),
    pct_diff = 100 * (auc - auc_ref) / auc_ref
  )
knitr::kable(dplyr::select(fig1c, Covariate, Dosing, pct_diff), digits = 1,
             caption = "Typical-subject AUCss difference (%) at the Figure 1C percentile thresholds.")

wb <- fig1c$pct_diff[fig1c$Dosing == "weight-based"]
fx <- fig1c$pct_diff[fig1c$Dosing == "fixed 71 mg"]
stopifnot(
  all(abs(wb) < 20),                                        # Results: < 20% for albumin and RBCT
  all(abs(wb[grepl("Weight", fig1c$Covariate[fig1c$Dosing == "weight-based"])]) < 10),  # < 10% for weight
  all(abs(fx) > 15)                                         # fixed dosing: about 25%
)
```

The thresholds are the edges of the extreme groups, so the group medians the
paper compares lie further out and its percentages (up to about 25% for fixed
dosing) are somewhat larger than these threshold values. The ordering --
weight-based dosing within 10% for weight, fixed dosing about 20% or more --
reproduces.

# Exposure-response

Each subject's exposure metric is the starting-dose AUCss = dose / individual
CL/F, taken from the PK simulation above. The 1 mg/kg arm is used because
79.8% of the PK population started at that dose.

```{r er-setup}
indiv <- sim |>
  dplyr::distinct(id, cl) |>
  dplyr::left_join(dplyr::select(cohort, id, dose_mg, treatment), by = "id") |>
  dplyr::mutate(AUC_LUSP = dose_mg / cl)

er_names <- c("Chen_2021_luspatercept_hb_change", "Chen_2021_luspatercept_hb_response",
              "Chen_2021_luspatercept_teae_grade3", "Chen_2021_luspatercept_bone_pain")
er <- lapply(er_names, function(n) rxode2::rxode2(readModelDb(n)))
names(er) <- er_names

predict_er <- function(ui, auc, out) {
  ev <- as.data.frame(rxode2::et(seq_along(auc) - 1))
  ev$AUC_LUSP <- auc
  res <- rxode2::rxSolve(ui, ev, returnType = "data.frame")  # no random effects in these models
  res[[out]]
}
auc_grid <- seq(0, 300, by = 5)
curves <- dplyr::bind_rows(
  tibble::tibble(endpoint = "Hb change, weeks 1-3 (g/dL)", auc = auc_grid,
                 value = predict_er(er[[1]], auc_grid, "dhb")),
  tibble::tibble(endpoint = "P(Hb increase >= 1 g/dL, day 21)", auc = auc_grid,
                 value = predict_er(er[[2]], auc_grid, "prob_hb_increase_1g")),
  tibble::tibble(endpoint = "P(TEAE grade >= 3, cycles 1-2)", auc = auc_grid,
                 value = predict_er(er[[3]], auc_grid, "prob_teae_grade3")),
  tibble::tibble(endpoint = "P(bone pain, cycles 1-2)", auc = auc_grid,
                 value = predict_er(er[[4]], auc_grid, "prob_bone_pain"))
)
```

## Figure 2A -- hemoglobin change over the first cycle

```{r figure-2a}
ggplot(fig2a, aes(auc, dhb)) +
  geom_point(aes(colour = dose)) +
  geom_line(data = dplyr::filter(curves, endpoint == "Hb change, weeks 1-3 (g/dL)", auc <= 215),
            aes(auc, value)) +
  labs(x = "Luspatercept AUCss (ug*day/mL)", y = "Hb change (g/dL/3 weeks)", colour = "mg/kg",
       caption = "Replicates Chen 2021 Figure 2A: digitized points and the packaged regression line.") +
  theme_minimal()

dhb_129 <- predict_er(er[[1]], 129, "dhb")
stopifnot(abs(dhb_129 - 1) < 0.15)  # Results: 'a mean increase of approximately 1 g/dL ... at 1 mg/kg (mean AUCss = 129)'
```

At the published mean 1 mg/kg exposure of 129 ug*day/mL the model predicts a
`r sprintf("%.2f", dhb_129)` g/dL increase, matching the Results statement of
"approximately 1 g/dL".

## Figure 2B -- probability of a >= 1 g/dL hemoglobin increase

The observed responder fractions by AUCss quartile (1/18, 4/17, 5/17, 10/18)
are compared with the model at each quartile's plotted position (read from
the vector coordinates of the Figure 2B markers).

```{r figure-2b}
q2b <- tibble::tibble(
  quartile = paste0("Q", 1:4),
  auc = x2b(c(360.38, 401.85, 419.40, 458.96)),
  observed = c(1 / 18, 4 / 17, 5 / 17, 10 / 18)
) |>
  dplyr::mutate(predicted = predict_er(er[[2]], auc, "prob_hb_increase_1g"))
knitr::kable(q2b, digits = 3)

or50 <- exp(50 * er[[2]]$iniDf$est[er[[2]]$iniDf$name == "e_auc_logit"])
stopifnot(
  abs(or50 - 3.73) < 0.01,
  max(abs(q2b$predicted - q2b$observed)) < 0.06
)

ggplot(dplyr::filter(curves, endpoint == "P(Hb increase >= 1 g/dL, day 21)", auc <= 215), aes(auc, value)) +
  geom_line() +
  geom_point(data = q2b, aes(auc, observed), shape = 15) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Luspatercept AUCss (ug*day/mL)", y = "Probability of response",
       caption = "Replicates Chen 2021 Figure 2B (squares: observed quartile fractions).") +
  theme_minimal()
```

## Figures 4A and 4B -- adverse events in cycles 1-2

A property of any maximum-likelihood logistic regression with an intercept
is that the fitted probabilities sum to the observed number of events. The
individual exposures are not published, but the simulated 1 mg/kg cohort
gives a comparable exposure distribution, so its mean predicted probability
should sit close to the observed overall rate (23/285 for grade >= 3 TEAEs,
55/285 for bone pain).

```{r figure-4}
auc_1mgkg <- indiv$AUC_LUSP[indiv$treatment == "1 mg/kg q3w"]
fig4 <- tibble::tibble(
  Endpoint = c("TEAE grade >= 3", "Bone pain (any grade)"),
  `Observed rate` = c(23 / 285, 55 / 285),
  `Mean predicted` = c(mean(predict_er(er[[3]], auc_1mgkg, "prob_teae_grade3")),
                       mean(predict_er(er[[4]], auc_1mgkg, "prob_bone_pain")))
)
knitr::kable(fig4, digits = 3)
stopifnot(abs(fig4$`Mean predicted` / fig4$`Observed rate` - 1) < 0.1)

curves |>
  dplyr::filter(grepl("cycles 1-2", endpoint), auc >= 13) |>
  ggplot(aes(auc, value)) +
  geom_line() +
  facet_wrap(~endpoint) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Luspatercept AUCss (ug*day/mL)", y = "Probability of event",
       caption = "Replicates the fitted lines of Chen 2021 Figures 4A and 4B.") +
  theme_minimal()
```

Both adverse-event relationships are flat, as the paper reports: across the
plotted exposure range the probability of a grade >= 3 TEAE changes by less
than half a percentage point and the probability of bone pain falls slightly
with exposure.

# Assumptions and deviations

* **IIV scale.** Table 2 reports IIV as percentages for an exponential model.
  They are read as sqrt(omega^2) x 100 (omega^2 = 0.1204 for CL/F, 0.0762 for
  V1/F), the convention of the same group's Chen 2020 luspatercept analysis.
  The alternative reading, omega^2 = log(1 + CV^2), would give 0.1135 and
  0.0735. The 36% descriptive CV of AUCss is consistent with either, so the
  choice has little practical effect.
* **Residual error.** The additive error on log-transformed concentrations
  is encoded as a proportional error of 20.8% in linear space.
* **Covariate centring.** The transfusion-burden terms are centred at 14
  units/24 weeks exactly as printed in the Results equations (the Table 1
  median is 14.1).
* **Figure-derived exposure-response coefficients.** Chen 2021 prints only
  one coefficient of the Figure 2A and 2B regressions (the slope and the odds
  ratio) and no coefficients for Figures 4A and 4B. The missing coefficients
  were read from the vector coordinates of the fitted lines, as shown above.
  Where a slope is printed, the printed value is used and the intercept is
  chosen so the line passes through the drawn line's midpoint (Figure 2A) or
  middle curve node (Figure 2B). The residual SD of the Figure 2A regression
  is not printed; 0.511 g/dL is the residual standard error of an OLS refit
  to the digitized points.
* **Placeholder residuals.** The three logistic models carry a fixed 0.001
  additive residual on the probability so that rxode2 has an error model to
  attach. The source likelihood is Bernoulli and has no residual error.
* **Extrapolated intercepts.** Placebo patients were shown in Figures 2 and 4
  for comparison but excluded from the regressions, so every intercept is an
  extrapolation to zero exposure rather than a placebo-arm estimate.
* **Analyses not packaged.** The multivariable logistic models for the phase 3
  transfusion-reduction endpoints (Table 3) do not report intercepts, and
  their exposure terms are not significant. The univariate adverse-event
  logistic models over the whole study (Supplemental Table S3) also omit the
  intercepts. The Cox models (Table 4) omit the baseline hazard. The
  dose-escalation-stratified curves of Figure 2D are a diagnostic for
  selection bias; the paper pools the strata and reports the pooled endpoint
  in Table 3. None of these can be simulated from the published information.
* **Virtual cohort.** Covariate distributions are assumed (log-normal weight,
  normal albumin and transfusion burden) and matched to the Table 1 medians
  and ranges; the individual data are not available.
* **Errata.** No erratum or correction for Chen 2021 was found in a
  EuropePMC search on 2026-09-27.
