---
title: "Methotrexate monotherapy in rheumatoid arthritis: DAS28, ACR20 and ACR50 MBMA (Chen 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methotrexate monotherapy in rheumatoid arthritis: DAS28, ACR20 and ACR50 MBMA (Chen 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7.5, fig.height = 5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Chen S, Wu Y, Huang W, Zhou J, Wei Z, Wu X. *Can Methotrexate Monotherapy
Achieve Clinical Remission in Patients with Active Rheumatoid Arthritis? A
Model-Based Meta-Analysis.* J Clin Pharmacol. 2025;65(10):1310-1321.
[doi:10.1002/jcph.70039](https://doi.org/10.1002/jcph.70039) (PMC12484399,
open access).

The authors performed a systematic search of PubMed, Cochrane and EMBASE for
randomised controlled trials of methotrexate (MTX) monotherapy in rheumatoid
arthritis (RA), pooled 69 trials covering 7999 patients, and fitted **three
independent model-based meta-analysis (MBMA) time-course models** -- one per
efficacy endpoint. Models for HAQ and ACR70 were attempted but could not be
fitted because of missing data.

Each endpoint is described by the same three-parameter Emax-in-time model
(Chen 2025 Methods, *Model Developing*):

$$E_{i,j} = \frac{E_{max} \cdot e^{\eta_i} \cdot Time_j}{ET_{50} + Time_j}$$

The single study-level random effect `eta_i` multiplies `Emax` only. The
display equation *as typeset* in the paper puts the same `exp(eta_i)` on `ET50`
as well; that is a typesetting artefact, and the reasoning is set out under
Assumptions and deviations below. Because the three endpoints were fitted
separately -- on different subsets of trials (31, 40 and 41 respectively), with
their own `Emax`, `ET50`, inter-study variance and residual error, and their own
bootstrap -- they are extracted as three model files that share this vignette:

```{r load-models}
# readModelDb() returns the model *function*; rxode2() gives the rxUi object
# whose $theta and $meta accessors are used below.
mod_das28 <- rxode2::rxode2(readModelDb("Chen_2025_methotrexate_das28_mbma"))
mod_acr20 <- rxode2::rxode2(readModelDb("Chen_2025_methotrexate_acr20_mbma"))
mod_acr50 <- rxode2::rxode2(readModelDb("Chen_2025_methotrexate_acr50_mbma"))

models <- list(DAS28 = mod_das28, ACR20 = mod_acr20, ACR50 = mod_acr50)
```

In all three models the observation `Cc` follows the package's canonical
observation name but is **not a drug concentration**: it is a study-arm summary
percentage. For DAS28 it is the percentage *change from baseline* (a negative
quantity); for ACR20 and ACR50 it is the percentage of patients in the arm
meeting the responder criterion (0-100).

# Population

```{r population}
pop <- mod_das28$meta$population
knitr::kable(
  data.frame(
    Field = c("Species", "Pooled patients", "Trials (DAS28 / ACR20 / ACR50)",
              "Disease state", "Baseline (pooled medians)",
              "Treatment duration"),
    Value = c(
      pop$species,
      format(pop$n_subjects, big.mark = ","),
      paste(mod_das28$meta$population$n_studies,
            mod_acr20$meta$population$n_studies,
            mod_acr50$meta$population$n_studies, sep = " / "),
      pop$disease_state,
      pop$baseline,
      "median 52.73 weeks (range 12-144)"
    )
  ),
  caption = "Chen 2025 Table 1 and Results, Characteristics of Included Studies."
)
```

Every modelled data point is one trial arm's mean endpoint value at one
follow-up time. Variability is **between-study**, not between-subject; these
models simulate study-arm mean trajectories and must not be used for
individual-patient prediction.

# Source trace

Every value in `ini()` and every equation in `model()`, with its location in the
source.

```{r source-trace, echo = FALSE}
knitr::kable(
  tibble::tribble(
    ~Item, ~`Source location`, ~Value,
    "Structural equation E = Emax*exp(eta)*t / (ET50 + t)", "Methods, *Model Developing*, display equation (PDF p. 1312) + Table 2 parameter labels", "eta on Emax only; printed equation also puts exp(eta) on ET50 (see Assumptions)",
    "Residual form Y = E + W*(ERR1 + RUV) / E*(1 + W*(ERR1+RUV))", "Methods, *Model Developing*, additive / proportional / exponential display equations", "form selected per endpoint (see Errata)",
    "Weight W = 1/sqrt(N)", "Methods, *Model Developing*, display equation", "N = per-arm sample size",
    "ERR1 realised once per study arm (NONMEM L2)", "Methods, *Model Developing*, prose", "encoded as eta_study_err1",
    "emax (DAS28)", "Table 2, Emax-DAS28", "-54.90 % (RSE 13%)",
    "let50 (DAS28)", "Table 2, ET50-DAS28", "log(20.60) week (RSE 26%)",
    "eta_study_emax (DAS28)", "Table 2, eta(Emax-DAS28) = 37.40%", "variance 0.374^2 = 0.139876",
    "eta_study_err1 (DAS28)", "Table 2, Correlation coefficient of DAS28", "0.00264 (RSE 28%)",
    "propSd (DAS28)", "Table 2, epsilon-DAS28 = 0.00135", "sqrt(0.00135) = 0.0367423",
    "emax (ACR20)", "Table 2, Emax-ACR20", "70.30 % (RSE 3%)",
    "let50 (ACR20)", "Table 2, ET50-ACR20", "log(6.69) week (RSE 7%)",
    "eta_study_emax (ACR20)", "Table 2, eta(Emax-ACR20) = 8.90%", "variance 0.089^2 = 0.007921",
    "eta_study_err1 (ACR20)", "Table 2, Correlation coefficient of ACR20", "1.98 (RSE 56%)",
    "addSd (ACR20)", "Table 2, epsilon-ACR20 = 1.91", "sqrt(1.91) = 1.3820275",
    "emax (ACR50)", "Table 2, Emax-ACR50", "49.4 % (RSE 24%)",
    "let50 (ACR50)", "Table 2, ET50-ACR50", "log(27.3) week (RSE 37%)",
    "eta_study_emax (ACR50)", "Table 2, eta(Emax-ACR50) = 45.40%", "variance 0.454^2 = 0.206116",
    "eta_study_err1 (ACR50)", "Table 2, Correlation coefficient of ACR50", "3.11 (RSE 106%)",
    "addSd (ACR50)", "Table 2, epsilon-ACR50 = 11.8", "sqrt(11.8) = 3.4351128",
    "Screened, NOT retained: MTX dose, RA duration, CRP, ESR", "Results, *Model Developing*; Tables S2-S6", "no covariate retained"
  ),
  caption = "Source trace for all three Chen 2025 endpoint models."
)
```

Two notes on how Table 2 was read:

* The `eta` rows are printed as percentages and the Methods state that ISV used
  an **exponential** model, so `eta = 37.40%` means `omega = 0.374` on the
  exponential scale. `ini()` carries the variance `omega^2`.
* The `Correlation coefficient` and `epsilon` rows both carry a **shrinkage**
  figure, which identifies them as NONMEM `$OMEGA` / `$SIGMA` **variances**
  rather than standard deviations. `ini()` therefore carries `sqrt(value)` for
  the residual SD.

# Errata

No erratum or corrigendum was located for Chen 2025.

The Supplemental Information (Table S1, the study-level overview; Tables S2-S6,
the covariate stepwise results; Figures S1-S6, risk-of-bias, goodness-of-fit and
LOO-CV plots) is referenced by the paper but was **not on disk** for this
extraction. No parameter value depends on it: the final models retained no
covariates, so every estimate is in Table 2 of the main text. Two consequences
are recorded rather than resolved:

1. **Study-count discrepancy.** The Results say "Table S1 provides an overview
   of the 71 studies" while the Abstract and the same Results paragraph both
   state 69 included studies. The model metadata uses 69.
2. **Per-endpoint patient counts are unavailable.** `n_subjects` records the
   pooled 7999 across all trials, not the per-endpoint subset.

**Internal inconsistency in the paper's own 12-week simulation numbers.** The
Results report the 12-week values as -20.3% (DAS28), 44.9% (ACR20) and 14.8%
(ACR50); the Discussion reports 21.17%, 46.84% and 17.63% for the same three
quantities at the same time point. The Results values are the ones reproduced
by the typical-value (eta = 0) model, as the regression test below shows, so
this vignette gates against the Results values and records the Discussion
values as unexplained.

This discrepancy is load-bearing for one of the paper's own conclusions. The
Discussion infers a EULAR "good response" at 12 weeks from a DAS28 unit change
of 1.23, computed as 21.17% of a 5.8 baseline. Recomputed from the Results value
of -20.3% against the Table 1 median baseline of 5.78, the change is 1.17 units
-- just below the 1.2-unit good-response threshold. The model reproduces the
Results value, so it reproduces 1.17, not 1.23. See the *Clinical interpretation
reproduced* section; no parameter was adjusted to move this number.

**Residual-error form was inferred, not stated.** The Methods list four
candidate residual forms (additive, proportional, exponential, combined) and
never say which was selected for which endpoint. The magnitudes in Table 2
settle it, because all three endpoints are modelled on the same percentage
scale:

| Endpoint | epsilon (variance) | as additive SD | as proportional CV | Verdict |
|---|---|---|---|---|
| DAS28 | 0.00135 | 0.037 pct pts on a -20% prediction | 3.7% | proportional |
| ACR20 | 1.91 | 1.38 pct pts on a 45-70% prediction | 138% | additive |
| ACR50 | 11.8 | 3.44 pct pts on a 15-23% prediction | 344% | additive |

An additive SD of 0.037 percentage points is impossible for arm means digitised
with WebPlotDigitizer (digitisation error alone is on the order of 1%), and a
proportional CV of 138% or 344% is impossible against the ACR predictions. The
files therefore encode DAS28 as proportional and ACR20/ACR50 as additive. A
per-endpoint selection is consistent with the paper's wording that the four
forms were candidates "with adjustments made to the model structures".

The alternative reading -- that DAS28 was fitted on the **fractional** scale
(`Emax = -0.549`) with an additive SD of 0.037, i.e. 3.7 percentage points after
rescaling -- cannot be excluded from the text. It would change only the DAS28
residual magnitude, not the structural model or any typical-value prediction.

**The paper's printed proportional and exponential residual equations are
dimensionally broken.** As typeset they read `Y = E * W * (1 + ERR1 + RUV)` and
`Y = E * W * exp(ERR1 + RUV)`, i.e. the weight multiplies the *prediction*,
which would scale every predicted value by roughly 0.1. The additive form
`Y = E + W * (ERR1 + RUV)` is dimensionally sound and is the form the weight is
clearly intended to take. These files place `W` on the error term only.

**Sample-size weighting is not applied inside the model.** The `propSd` /
`addSd` values are the unweighted Table 2 estimates. Chen 2025 weights each
study-arm observation by `W = 1/sqrt(N)`; downstream simulation code that wants
per-arm realism must scale the residual SD by `1/sqrt(N_arm)`. This follows the
`Boucher_2018_naproxen_mbma` and `Mercier_2014_tramadol_tapentadol_mbma`
precedent.

# PKNCA not applicable

These MBMA models have no drug-concentration output, no dose events and no
absorption-distribution-elimination profile to integrate -- MTX exposure does
not enter the model at all (dose was screened as a covariate and rejected).
PKNCA-style NCA (Cmax / Tmax / AUC / half-life) is therefore not a meaningful
validation target. The models are validated instead by (a) closed-form
structural identities, (b) a six-point regression test against the paper's
published 12- and 24-week simulation results, and (c) replication of the
published time-course figures -- the strategy used for
`Boucher_2018_naproxen_mbma` and `Vargo_2014_statins_ezetimibe_mbma`.

# Validation 1: closed-form structural identities

The Emax-in-time form implies three exact identities, checked here against the
typical-value (random-effects-zeroed) solve. Both sides use the same fixed
parameters, so the only difference is numerical and a tight bound is correct.

```{r structural-gates}
typ <- lapply(models, rxode2::zeroRe)

solve_typ <- function(m, times) {
  ev <- as.data.frame(rxode2::et(times))
  ev$id <- 1L
  rxode2::rxSolve(m, events = ev, returnType = "data.frame")$Cc
}

gates <- lapply(names(models), function(nm) {
  m     <- typ[[nm]]
  emax  <- models[[nm]]$theta[["emax"]]
  et50  <- exp(models[[nm]]$theta[["let50"]])
  # t = 0 -> 0; t = ET50 -> Emax/2; t -> large -> approaches Emax
  y <- solve_typ(m, c(0, et50, 1e6))
  data.frame(
    Endpoint     = nm,
    `At t=0`     = y[1],
    `At t=ET50`  = y[2],
    `Emax/2`     = emax / 2,
    `At t=1e6`   = y[3],
    Emax         = emax,
    check.names  = FALSE
  )
}) |> dplyr::bind_rows()

knitr::kable(gates, digits = 4,
             caption = "Closed-form identities of the Emax-in-time model.")

stopifnot(
  # E(0) = 0 exactly for every endpoint.
  all(abs(gates$`At t=0`) < 1e-8),
  # E(ET50) = Emax/2 exactly.
  all(abs(gates$`At t=ET50` - gates$`Emax/2`) < 1e-6),
  # E(t) -> Emax as t grows large.
  all(abs(gates$`At t=1e6` - gates$Emax) < 1e-3 * abs(gates$Emax))
)
```

# Validation 2: regression test against the published simulation results

Chen 2025 Results, *Model Simulation*: "MTX monotherapy resulted in DAS28
decline rates, ACR20, and ACR50 of -20.3%, 44.9%, and 14.8%, respectively, at 12
weeks, and -29.6%, 54.7%, and 22.7%, respectively, at 24 weeks."

Because the single `eta` enters monotonically, the median over the between-study
distribution equals the typical value at `eta = 0`, so those six published
medians are directly comparable to the typical-value solve. This is the
strongest available check on the structural form: it independently confirms
there is **no Hill exponent** (the Methods call the model "sigmoid Emax" but no
gamma is tabulated), and it confirms the `Emax` and `ET50` values and units.

```{r published-simulation-comparison}
published <- tibble::tribble(
  ~Endpoint, ~Week, ~Published,
  "DAS28", 12, -20.3,
  "DAS28", 24, -29.6,
  "ACR20", 12,  44.9,
  "ACR20", 24,  54.7,
  "ACR50", 12,  14.8,
  "ACR50", 24,  22.7
)

published$Model <- vapply(seq_len(nrow(published)), function(i) {
  solve_typ(typ[[published$Endpoint[i]]], c(0, published$Week[i]))[2]
}, numeric(1))

published <- published |>
  dplyr::mutate(`Difference (pct pts)` = Model - Published,
                `Relative difference (%)` = 100 * (Model - Published) / abs(Published))

knitr::kable(
  published |>
    dplyr::rename("Simulated (this package)" = Model,
                  "Published (Chen 2025)"    = Published),
  digits = 3,
  caption = "Chen 2025 Results, Model Simulation: published 12- and 24-week values vs the typical-value model."
)

stopifnot(
  # Deterministic typical-value solve against six published point estimates.
  # The residual gap is the paper's Monte Carlo / rounding noise only.
  max(abs(published$`Relative difference (%)`)) < 3,
  # Signs must match: DAS28 declines, ACR responder rates rise.
  all(sign(published$Model) == sign(published$Published))
)
```

Every published value is reproduced to better than 2% relative difference,
without tuning any parameter.

# Replication: published time-course figures (Chen 2025 Figures 3, 4, 5)

Figures 3, 4 and 5 plot the simulated time-varying DAS28 change rate, ACR20 and
ACR50 over the 52-week horizon the authors simulated, with the typical efficacy
as a dotted line and a shaded 95% interval. The panel below reproduces the
typical trajectories and the between-study interval.

```{r figures-3-4-5, fig.height = 4.2, fig.cap = "Replication of Chen 2025 Figures 3-5: simulated time course of DAS28 percentage change, ACR20 and ACR50 under MTX monotherapy over 52 weeks. Line = typical value (eta = 0); band = 5th-95th percentile across 200 simulated study arms."}
n_arms <- 200L                       # study arms per endpoint; <= 200 per the cohort cap
tgrid  <- seq(0, 52, by = 1)

rxode2::rxSetSeed(1234)

sim_one <- function(nm) {
  ev <- as.data.frame(rxode2::et(tgrid, id = seq_len(n_arms)))
  s  <- rxode2::rxSolve(models[[nm]], events = ev) |> as.data.frame()
  s$Endpoint <- nm
  s
}

sim_all <- dplyr::bind_rows(lapply(names(models), sim_one))

band <- sim_all |>
  dplyr::group_by(Endpoint, time) |>
  dplyr::summarise(lo  = quantile(Cc, 0.05),
                   mid = median(Cc),
                   hi  = quantile(Cc, 0.95),
                   .groups = "drop")

typical <- dplyr::bind_rows(lapply(names(models), function(nm) {
  data.frame(Endpoint = nm, time = tgrid, Cc = solve_typ(typ[[nm]], tgrid))
}))

ggplot(band, aes(x = time)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "steelblue", alpha = 0.25) +
  geom_line(data = typical, aes(y = Cc), linetype = "dotted", linewidth = 0.9) +
  facet_wrap(~ Endpoint, scales = "free_y") +
  labs(x = "Time (weeks)", y = "Percent",
       title = "Chen 2025 Figures 3-5 -- simulated MTX monotherapy time course",
       caption = "Dotted = typical value; band = 5th-95th percentile across 200 study arms.") +
  theme_bw()
```

The between-study band is widest for ACR50 (ISV 45.4%) and narrowest for ACR20
(ISV 8.9%), matching the ordering of the `eta(Emax)` estimates in Table 2.

# Validation 3: the median of the between-study cohort equals the typical value

Because `eta` enters `E(t)` monotonically, the cohort median must coincide with
the typical-value curve. This check does **not** discriminate between the
`eta`-on-`Emax`-only encoding used here and the `eta`-on-both reading of the
printed equation -- both are monotone in `eta` -- but a sign or scale error in
the `eta` encoding would break it.

The comparison is made on a robust central statistic rather than on extremes,
because the cohort draw is not reproducible across rxode2 versions.

```{r median-vs-typical}
med_chk <- band |>
  dplyr::inner_join(typical, by = c("Endpoint", "time")) |>
  dplyr::filter(time > 0) |>
  dplyr::mutate(pct_diff = 100 * (mid - Cc) / abs(Cc))

med_summary <- med_chk |>
  dplyr::group_by(Endpoint) |>
  dplyr::summarise(`Median abs % difference` = median(abs(pct_diff)),
                   `90th pctile abs % difference` = quantile(abs(pct_diff), 0.9),
                   .groups = "drop")

knitr::kable(med_summary, digits = 3,
             caption = "Cohort median vs typical value across the 52-week grid.")

stopifnot(
  # Centre: a scale or sign error in the eta encoding moves this by tens of pct.
  all(med_summary$`Median abs % difference` < 5),
  # Envelope: robust to which arms land in the tails.
  all(med_summary$`90th pctile abs % difference` < 15)
)
```

# Clinical interpretation reproduced

The paper's central claim is that MTX monotherapy does not reach remission. Two
statements are checkable against the model.

```{r clinical-claims}
# Discussion: "the DAS28 change rate reached 21.17% at 12 weeks, corresponding
# to a DAS28 score change of 1.23 (with an average baseline DAS28 score of 5.8
# ...)". A EULAR "good response" needs a drop > 1.2 units; "remission" needs a
# > 50% change rate.
das28_12wk_pct <- published$Model[published$Endpoint == "DAS28" &
                                    published$Week == 12]

claims <- tibble::tribble(
  ~Basis,                            ~`Change rate (%)`, ~`Baseline DAS28`,
  "Model (reproduces paper Results)", abs(das28_12wk_pct), 5.78,
  "Paper Discussion arithmetic",      21.17,               5.80
) |>
  dplyr::mutate(
    `Unit change` = `Change rate (%)` / 100 * `Baseline DAS28`,
    `Good response (> 1.2 units)` = ifelse(`Unit change` > 1.2, "yes", "no"),
    `Remission (> 50% change)`    = ifelse(`Change rate (%)` > 50, "yes", "no")
  )

knitr::kable(claims, digits = 3,
             caption = "Chen 2025 Discussion, DAS28 paragraph: the good-response claim recomputed from the Results value and from the Discussion's own value.")

stopifnot(
  # The paper's CENTRAL conclusion -- the 50% change rate required for clinical
  # remission is not reached at 12 weeks -- holds on either basis. This is the
  # robust, directional claim and is the one gated here.
  all(claims$`Change rate (%)` < 50),
  # The model reproduces the paper's Results 12-week value (gated exactly in
  # the regression test above); the unit change it implies is ~1.17.
  abs(claims$`Unit change`[1] - 1.168) < 0.02
)
```

The model reproduces the paper's central conclusion -- at 12 weeks MTX
monotherapy falls well short of the 50% DAS28 change rate that would indicate
clinical remission -- on either arithmetic basis.

The *secondary* claim, that a EULAR good response (a drop of more than 1.2
DAS28 units) is reached at 12 weeks, does **not** survive recomputation from the
paper's own Results. It depends entirely on which of the paper's two mutually
inconsistent 12-week figures is used: the Discussion's 21.17% against a 5.8
baseline gives 1.23 units and clears the threshold, whereas the Results' -20.3%
-- the value the model reproduces to 0.5% -- gives 1.17 units against the Table 1
median baseline of 5.78 and falls just short. This is a reporting inconsistency
in the source, not an encoding error: the model matches the Results value
exactly, and no parameter was adjusted. It is recorded here and in Errata rather
than gated.

# Assumptions and deviations

* **`eta` on `Emax` only, against the printed equation.** The display equation
  in *Model Developing* is typeset as
  `E = Emax*exp(eta_i)*t / (ET50*exp(eta_i) + t)`, placing the same random
  effect on `ET50` as well as on `Emax`. These files deliberately depart from
  that and put `eta` on `Emax` alone, because three independent statements in
  the paper contradict the typeset form:

    1. Table 2 names the parameter `eta (Emax-DAS28)` / `eta (Emax-ACR20)` /
       `eta (Emax-ACR50)` and reports exactly **one** random effect per
       endpoint -- there is no `eta(ET50)` row anywhere in the table.
    2. The Results state that "Covariates were not investigated on ET50
       because ISV was unsuccessful in being evaluated", i.e. inter-study
       variability on `ET50` was not estimable at all.
    3. The Results describe the estimates as "the ISV for `Emax` is 37.4%,
       8.9%, and 45.4%".

  The usual "trust the printed equation over the prose" rule is not applied
  here because this paper's display equations are demonstrably unreliable --
  its residual-error equations are dimensionally broken as typeset (see
  Errata), so the typesetting is not trustworthy evidence against three
  consistent textual and tabular statements. **No typical-value prediction
  changes** under either reading (`eta = 0` gives `exp(eta) = 1`), so every
  regression test against the paper's published simulation values below is
  unaffected; only the width and shape of the between-study band differ.
  Users who want the equation-as-printed reading can multiply `et50Arm` by
  `exp(eta_study_emax)`.
* **No Hill exponent.** The Methods call the model "sigmoid Emax" but Table 2
  reports no gamma, and the gamma = 1 form reproduces all six published
  simulation values to better than 2%. Encoded with gamma = 1.
* **`eta_study_err1` has no paired fixed effect**, so
  `checkModelConventions()` emits one warning per file
  (`IIV 'eta_study_err1' has no matching fixed-effect parameter`). This is
  expected and is not a naming slip: `ERR1` is the paper's zero-mean,
  study-arm-level residual-correlation term (NONMEM L2 grouping), and a
  zero-mean random effect on the response has no fixed counterpart to pair
  with. The alternative -- folding `ERR1` into the residual as a combined
  variance -- would silence the warning but discard a parameter that the paper
  reports with its own RSE and bootstrap interval, so the faithful encoding is
  kept.
* **Residual form and sample-size weighting** are as described in Errata: the
  form was inferred from Table 2 magnitudes, and the `1/sqrt(N)` weight is left
  to downstream code.
* **ACR20 inter-study variability is weakly identified** (eta-shrinkage 49% per
  Table 2). The ACR20 between-study band above should be read with that caveat.
* **ACR50 is the weakest of the three models** overall: the authors note its
  goodness-of-fit trend lines were "not as parallel to the reference line", the
  correlation coefficient has a 106% RSE, and the bootstrap median `Emax`
  (54.6%) sits well above the point estimate (49.4%).
* **No covariates.** Every simulation here is covariate-free because the final
  models are. MTX dose, RA disease duration, baseline CRP and baseline ESR were
  screened and rejected; the first three are recorded in
  `covariatesDataExcluded`. Baseline ESR is recorded in prose in the model
  `population$notes` instead, because the covariate register has no canonical
  entry for erythrocyte sedimentation rate and these files do not introduce a
  new canonical name for a covariate that carries no coefficient.
* **Between-study, not between-subject.** `eta_study_emax` and
  `eta_study_err1` are study-arm-level effects. Simulated trajectories are
  study-arm means; individual-patient simulation is out of scope.
