---
title: "Tacrolimus (Chen 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tacrolimus (Chen 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Chen_2020_tacrolimus"))$meta$reference`
* Description: `r rxode2::rxode(readModelDb("Chen_2020_tacrolimus"))$meta$description`
* Article: <https://doi.org/10.3892/etm.2020.8821> (open access, PMC7388563)

Chen et al. (2020) fitted a one-compartment model with first-order absorption
to routine therapeutic-drug-monitoring whole-blood concentrations of oral
tacrolimus in children and adolescents with lupus nephritis (LN), and used
Monte Carlo simulation to recommend weight-banded initial doses with and
without the CYP3A-inhibiting Chinese patent medicine Wuzhi capsule. The
absorption rate constant was fixed at 4.48 1/h from earlier pediatric
tacrolimus models (the paper's references 13 and 16-18).

# Population

Thirty-two Chinese pediatric and adolescent LN patients (5 male, 27 female)
treated at the Children's Hospital of Fudan University between August 2014 and
September 2019. From Table I: age median 13.87 (range 2.86-17.99) years, body
weight median 47.00 (17.00-66.50) kg. All patients received a glucocorticoid
and 12 of 32 received Wuzhi capsule, the only co-medication retained on CL/F.

# Source trace

| Element | Value | Source |
|---|---|---|
| Structure: 1-compartment, first-order absorption and elimination | -- | Methods, Population pharmacokinetics modeling |
| `lka` | 4.48 1/h (fixed) | Table II; Methods |
| `lcl` | 15.5 L/h | Table II; Results equation vi |
| `lvc` | 174 L | Table II; Results equation vii |
| `e_wt_cl` | 0.75 (fixed), reference 70 kg | Methods equation iii; equation vi |
| `e_wt_vc` | 1 (fixed), reference 70 kg | Methods equation iii; equation vii |
| `e_conmed_wuzhi_cl` | -0.290 | Table II (theta WZ); equation vi |
| Wuzhi form `(1 + WZ * theta)` | -- | Methods equation v; equation vi |
| `etalcl` | 0.172^2 = 0.029584 | Table II (omega CL/F = 0.172, read as SD; see below) |
| Exponential IIV `A = T(A) * exp(eta)` | -- | Methods equation i |
| `propSd` | 0.281 | Table II (sigma1, proportional); Methods equation ii |

# Apparent clearance by weight (Figure 3)

Figure 3 plots the typical weight-normalised CL/F with and without Wuzhi
capsule at the six simulated body weights. The values below were digitised by
the maintainers from Figure 3.

```{r fig3}
mod <- readModelDb("Chen_2020_tacrolimus")
mod_typ <- rxode2::zeroRe(mod)

fig3 <- tibble::tribble(
  ~WT, ~CONMED_WUZHI, ~pub_cl_kg,
  10, 0, 0.360, 20, 0, 0.302, 30, 0, 0.272,
  40, 0, 0.253, 50, 0, 0.240, 60, 0, 0.230,
  10, 1, 0.254, 20, 1, 0.214, 30, 1, 0.193,
  40, 1, 0.180, 50, 1, 0.170, 60, 1, 0.163
)

ev3 <- fig3 |>
  mutate(id = row_number(), time = 0, amt = 0, evid = 0L, cmt = "central") |>
  select(id, time, amt, evid, cmt, WT, CONMED_WUZHI)
cl3 <- rxode2::rxSolve(mod_typ, events = ev3, keep = c("WT", "CONMED_WUZHI")) |>
  as.data.frame() |>
  select(WT, CONMED_WUZHI, cl)

cmp3 <- inner_join(fig3, cl3, by = c("WT", "CONMED_WUZHI")) |>
  mutate(sim_cl_kg = cl / WT, pct_diff = 100 * (sim_cl_kg - pub_cl_kg) / pub_cl_kg)

cmp3 |>
  select(WT, CONMED_WUZHI, pub_cl_kg, sim_cl_kg, pct_diff) |>
  dplyr::rename(
    "Weight (kg)" = WT, "Wuzhi capsule" = CONMED_WUZHI,
    "Figure 3 CL/F (L/h/kg)" = pub_cl_kg, "Model CL/F (L/h/kg)" = sim_cl_kg,
    "% diff" = pct_diff
  ) |>
  knitr::kable(digits = 3, caption = "Replicates Figure 3 of Chen 2020.")

# Typical values against a digitised curve: the only difference is reading
# error off the figure.
stopifnot(all(abs(cmp3$pct_diff) < 3))
```

# Probability of target attainment (Figure 4)

The paper simulates six body weights (10-60 kg) and seven initial regimens
(0.01-0.30 mg/kg/day split into two doses) with and without Wuzhi capsule, and
plots the probability that the concentration falls in the 5-15 ng/mL window.
We replicate it with 200 subjects per weight-by-dose-by-Wuzhi arm (the package
cap), simulating 7 days of twice-daily dosing and reading the steady-state
trough at the end of the last dosing interval. The model's `Cc` output is the
individual prediction (no residual error); this is what reproduces Figure 4
(see Assumptions). The published probabilities below were digitised by the
maintainers from Figure 4 at the six simulated weights.

```{r fig4-sim}
weights <- c(10, 20, 30, 40, 50, 60)
doses_kg <- c(0.01, 0.05, 0.10, 0.15, 0.20, 0.25, 0.30)
nsub <- 200
ndays <- 7
tobs <- ndays * 24

arms <- tidyr::expand_grid(CONMED_WUZHI = 0:1, dose_kg = doses_kg, WT = weights) |>
  mutate(arm = row_number())

subj <- arms |>
  tidyr::uncount(nsub) |>
  mutate(id = row_number())
dtimes <- seq(0, by = 12, length.out = 2 * ndays)

ev_dose <- subj |>
  tidyr::expand_grid(time = dtimes) |>
  mutate(amt = dose_kg * WT / 2, evid = 1L, cmt = "depot")
ev_obs <- subj |>
  mutate(time = tobs, amt = 0, evid = 0L, cmt = "central")
ev4 <- bind_rows(ev_dose, ev_obs) |>
  arrange(id, time, desc(evid)) |>
  select(id, time, amt, evid, cmt, WT, CONMED_WUZHI, arm)

rxode2::rxSetSeed(2020)
sim4 <- rxode2::rxSolve(mod, events = ev4, keep = "arm") |>
  as.data.frame() |>
  filter(time == tobs)

pta <- sim4 |>
  group_by(arm) |>
  summarise(sim_pta = 100 * mean(Cc >= 5 & Cc <= 15), .groups = "drop") |>
  left_join(arms, by = "arm")
```

```{r fig4-compare}
pub4 <- tibble::tribble(
  ~CONMED_WUZHI, ~dose_kg, ~p10, ~p20, ~p30, ~p40, ~p50, ~p60,
  0, 0.01, 0, 0, 0, 0, 0, 0,
  0, 0.05, 1, 9, 22, 36, 47, 56,
  0, 0.10, 45, 82, 91, 93, 93, 91,
  0, 0.15, 84, 88, 78, 65, 53, 44,
  0, 0.20, 85, 65, 44, 26, 17, 12,
  0, 0.25, 73, 37, 17, 10, 6, 4,
  0, 0.30, 55, 18, 9, 4, 2, 0,
  1, 0.01, 0, 0, 0, 0, 0, 0,
  1, 0.05, 34, 72, 88, 93, 96, 98,
  1, 0.10, 93, 85, 67, 53, 41, 31,
  1, 0.15, 66, 28, 12, 6, 4, 2,
  1, 0.20, 28, 6, 1, 0, 0, 0,
  1, 0.25, 11, 1, 0, 0, 0, 0,
  1, 0.30, 5, 0, 0, 0, 0, 0
) |>
  tidyr::pivot_longer(p10:p60, names_to = "WT", values_to = "pub_pta") |>
  mutate(WT = as.numeric(sub("p", "", WT)))

cmp4 <- inner_join(pub4, pta, by = c("CONMED_WUZHI", "dose_kg", "WT")) |>
  mutate(diff_pts = sim_pta - pub_pta)

cmp4 |>
  filter(dose_kg != 0.01) |>
  mutate(cell = sprintf("%.0f / %.0f", pub_pta, sim_pta)) |>
  select(CONMED_WUZHI, dose_kg, WT, cell) |>
  tidyr::pivot_wider(names_from = WT, values_from = cell, names_prefix = "wt_") |>
  dplyr::rename(
    "Wuzhi capsule" = CONMED_WUZHI, "Dose (mg/kg/day)" = dose_kg,
    "10 kg" = wt_10, "20 kg" = wt_20, "30 kg" = wt_30,
    "40 kg" = wt_40, "50 kg" = wt_50, "60 kg" = wt_60
  ) |>
  knitr::kable(caption = "Replicates Figure 4 of Chen 2020: probability (%) of a trough in 5-15 ng/mL, published / simulated.")
```

```{r fig4-assert}
# The probabilities depend jointly on the typical values, the Wuzhi effect and
# the IIV scale; a wrong omega reading (variance instead of SD) flattens every
# curve to 40-60% and fails both bounds by a wide margin. Bounds are on the
# centre and the 90th percentile of the absolute difference (percentage
# points), robust to Monte Carlo noise at n = 200 per arm and to digitisation.
stopifnot(
  median(abs(cmp4$diff_pts)) < 5,
  quantile(abs(cmp4$diff_pts), 0.9) < 15
)
```

```{r fig4-fig, fig.cap = "Simulated probability of a 5-15 ng/mL trough (lines) against values digitised from Figure 4 of Chen 2020 (points). Replicates Figure 4 of Chen 2020."}
cmp4 |>
  mutate(
    dose = factor(sprintf("%.2f mg/kg/day", dose_kg)),
    panel = ifelse(CONMED_WUZHI == 1, "(B) with Wuzhi capsule", "(A) without Wuzhi capsule")
  ) |>
  ggplot(aes(WT, colour = dose)) +
  geom_line(aes(y = sim_pta)) +
  geom_point(aes(y = pub_pta)) +
  facet_wrap(~panel) +
  labs(x = "Body weight (kg)", y = "Probability of target attainment (%)", colour = "Dose")
```

The recommended regimens of Table III follow from these curves: without Wuzhi
capsule 0.15 mg/kg/day below 23 kg and 0.10 mg/kg/day above; with Wuzhi capsule
0.10 and 0.05 mg/kg/day respectively.

# PKNCA validation

The paper reports no NCA. As a structural check, typical-value subjects
(`zeroRe()`) at each Figure 4 weight, with and without Wuzhi capsule, receive
0.10 mg/kg/day split into two doses for 7 days. PKNCA computes the
steady-state AUC over the last dosing interval, which must equal the closed
form `Dose / (CL/F)`.

```{r pknca}
tlast <- tobs - 12
ev_nca <- tidyr::expand_grid(CONMED_WUZHI = 0:1, WT = weights) |>
  mutate(id = row_number())
nca_dose <- ev_nca |>
  tidyr::expand_grid(time = dtimes) |>
  mutate(amt = 0.10 * WT / 2, evid = 1L, cmt = "depot")
nca_obs <- ev_nca |>
  tidyr::expand_grid(time = tlast + c(0, 0.25, 0.5, 1, 1.5, 2, 3, 4, 6, 8, 10, 12)) |>
  mutate(amt = 0, evid = 0L, cmt = "central")
ev_nca_all <- bind_rows(nca_dose, nca_obs) |>
  arrange(id, time, desc(evid)) |>
  mutate(treatment = paste0(WT, " kg, WZ=", CONMED_WUZHI))

sim_nca <- rxode2::rxSolve(mod_typ, events = ev_nca_all, keep = c("WT", "CONMED_WUZHI", "treatment")) |>
  as.data.frame()

conc <- sim_nca |>
  filter(!is.na(Cc), time >= tlast) |>
  select(id, time, Cc, treatment)
doses <- ev_nca_all |>
  filter(evid == 1, time == tlast) |>
  select(id, time, amt, treatment)

o_conc <- PKNCAconc(conc, Cc ~ time | treatment + id)
o_dose <- PKNCAdose(doses, amt ~ time | treatment + id)
intervals <- data.frame(start = tlast, end = tlast + 12, auclast = TRUE, cmin = TRUE, cmax = TRUE)
nca_res <- pk.nca(PKNCAdata(o_conc, o_dose, intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  left_join(distinct(ev_nca_all, treatment, WT, CONMED_WUZHI), by = "treatment") |>
  mutate(auc_closed = 1000 * 0.10 * WT / 2 / (15.5 * (WT / 70)^0.75 * (1 - 0.290 * CONMED_WUZHI))) |>
  arrange(CONMED_WUZHI, WT)

nca_wide |>
  select(WT, CONMED_WUZHI, auclast, auc_closed, cmax, cmin) |>
  dplyr::rename(
    "Weight (kg)" = WT, "Wuzhi capsule" = CONMED_WUZHI,
    "AUCtau (PKNCA, ng*h/mL)" = auclast, "Dose/(CL/F) (ng*h/mL)" = auc_closed,
    "Cmax (ng/mL)" = cmax, "Cmin (ng/mL)" = cmin
  ) |>
  knitr::kable(digits = 2, caption = "Typical-value steady-state NCA, 0.10 mg/kg/day in two doses.")

# Same typical parameters on both sides: the only difference is trapezoidal
# error over a smooth 12-point interval, so a tight bound applies.
stopifnot(all(abs(nca_wide$auclast / nca_wide$auc_closed - 1) < 0.02))
```

# Assumptions and deviations

* **IIV scale.** Table II reports "omega CL/F = 0.172" without stating whether
  it is a variance or an SD. Simulating Figure 4 under both readings settles
  it: with 0.172 as the SD of eta (variance 0.029584) the simulated
  probabilities follow the published curves (e.g. 40 kg, 0.10 mg/kg/day, no
  Wuzhi: about 93% simulated vs 93% published; 60 kg, 0.05 mg/kg/day with
  Wuzhi: about 97% vs 98%), whereas treating 0.172 as the variance (SD 0.415)
  flattens every curve to roughly 40-60% (about 56% and 64% at those two
  points). The SD reading is used, the same convention the sister paper from
  this centre (Wang 2019, SOJIA; `Wang_2019b_tacrolimus`) resolves to.
* **Residual error scale and Figure 4.** sigma1 = 0.281 is taken as the
  proportional SD (28.1%), consistent with the omega reading. Figure 4 is
  reproduced by individual predictions without residual error; adding the
  proportional error lowers the 40 kg, 0.10 mg/kg/day probability to about
  81%, well below the published 93%.
* **Sampling time.** The paper does not state when the TDM samples were
  drawn. The fixed ka and the Figure 4 replication (steady-state trough before
  the next dose) are consistent with trough sampling.
* **Figure 4 regimen.** "Split into two doses" is read as equal doses every
  12 h. The simulation runs 7 days (more than 10 half-lives at every weight)
  in place of an explicit steady-state solve.
* **Digitisation.** Figure 3 and Figure 4 values were read off the published
  figures at the six simulated weights; Excel-style smoothing between those
  points is ignored. The largest Figure 4 discrepancy (about 8 percentage
  points at 20 kg without Wuzhi) is at the steepest part of the curves.
* **Wuzhi capsule indicator.** The paper does not state whether it was
  time-varying; it is accepted per record. The Wuzhi dose is not reported.
* **Fixed ka** is inherited from earlier pediatric models and is not
  identifiable from TDM data.
* The body-weight reference (70 kg) lies above the observed median (47 kg);
  the typical CL/F of 15.5 L/h is an adult-equivalent value.
