---
title: "Ciclosporin (Chen 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ciclosporin (Chen 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Chen_2020_ciclosporin"))$meta$reference`
* Description: `r rxode2::rxode(readModelDb("Chen_2020_ciclosporin"))$meta$description`
* Article: <https://doi.org/10.3892/etm.2020.8732>

Chen et al. (2020) developed a population PK model for oral ciclosporin in
18 Chinese children who underwent bone marrow transplantation (BMT) at the
Children's Hospital of Fudan University between September 2016 and
September 2019, and used Monte Carlo simulation to choose an initial dose.
Every concentration was a whole-blood trough from routine therapeutic drug
monitoring, so the absorption rate constant was fixed at 0.68 1/h from the
literature (the paper's references 16, 21 and 25; reference 21 is the Ni
2013 paediatric model, `Ni_2013_ciclosporin`, and reference 25 is the same
group's `Wang_2019_cyclosporin`). The structural model is one-compartment
with first-order absorption and elimination, in apparent (F-scaled) terms.

The final covariate model (Chen 2020 Results, Equations F and G) is

$$
\mathrm{CL/F} = 29.2 \times \left(\frac{\mathrm{WT}}{70}\right)^{0.75}
  \times \left(\frac{\mathrm{POD}}{51.5}\right)^{0.749} \quad \mathrm{L/h}
$$

$$
\mathrm{V/F} = 6550 \times \frac{\mathrm{WT}}{70} \quad \mathrm{L}
$$

where POD is days post-transplant, normalised to the cohort median of 51.5
days (Table I). Apparent clearance rises with time after transplant.

# Population

Chen 2020 Table I: n = 18 children (13 male / 5 female; 27.8% female), age
1.60 +/- 1.15 years (median 1.22, range 0.29-6.49), body weight 8.40 +/-
3.28 kg (median 7.60, range 5.20-25.60), POD 61.16 +/- 40.16 days (median
51.50, range 1-188). Initial ciclosporin doses were 14-100 mg/day, later
adjusted on efficacy, adverse events and trough concentration. Co-medications:
glucocorticoids 15, omeprazole 16, mycophenolate mofetil 7, phenobarbital 2
and tacrolimus 2 of 18; none was retained as a covariate. The same
information is in the model's `population` metadata.

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` (Ka) | `fixed(log(0.68))` 1/h | Methods 'Population pharmacokinetic modeling'; Table II |
| `lcl` (CL/F at 70 kg, POD 51.5 d) | `log(29.2)` L/h | Table II; Results Equation F |
| `lvc` (V/F at 70 kg) | `log(6550)` L | Table II; Results Equation G |
| `e_wt_cl` | `fixed(0.75)` | Methods Equation C (reference 26) |
| `e_wt_vc` | `fixed(1)` | Methods Equation C (reference 26) |
| `e_pod_cl` | `0.749` | Table II 'theta POD'; Equation F |
| POD reference | 51.5 days | Equation F; Table I median |
| IIV CL/F | `etalcl ~ 0.627^2` | Table II 'omega CL/F' = 0.627, read as SD (see below) |
| IIV V/F | `etalvc ~ 0.998^2` | Table II 'omega V/F' = 0.998, read as SD (see below) |
| Proportional residual | `propSd = 0.447` | Table II 'sigma 1' |
| Additive residual | `addSd = 70.071` ng/mL | Table II 'sigma 2' |
| IIV model | `P_i = T(P) * exp(eta_i)` | Methods Equation A |
| Residual model | `Y = F * (1 + eps1) + eps2` | Methods Equation B |
| Units | `Cc <- central / vc * 1000` | mg / L = mg/L; x1000 gives the ng/mL of Tables II-III |

# Setup

```{r load-model}
mod <- readModelDb("Chen_2020_ciclosporin")
ui <- rxode2::rxode(mod)
mod_typical <- mod |> rxode2::zeroRe()
stopifnot(identical(ui$state, c("depot", "central")))

published_cl <- function(wt, pod) 29.2 * (wt / 70)^0.75 * (pod / 51.5)^0.749
published_vc <- function(wt) 6550 * wt / 70
```

# Structural gate: Figure 3 (CL/F per kg against POD)

Chen 2020 Figure 3 plots weight-normalised CL/F against POD for 5, 10, 20
and 30 kg children. The encoded model is evaluated at the right-hand end of
that figure (POD = 188 days, the largest observed) and compared with values
read from the figure, and with the published equation to solver precision.

```{r fig3-gate}
fig3 <- tibble::tibble(
  WT = c(5, 10, 20, 30),
  # Read by the maintainers from Chen 2020 Figure 3 at POD = 188 days.
  cl_per_kg_fig = c(2.13, 1.79, 1.51, 1.36)
)
pod_grid <- c(1, 10, 25, 51.5, 100, 150, 188)
grid <- tidyr::expand_grid(WT = fig3$WT, POD = pod_grid) |>
  dplyr::mutate(id = dplyr::row_number())

ev_grid <- dplyr::bind_rows(lapply(seq_len(nrow(grid)), function(i) {
  tibble::tibble(
    id = grid$id[i], time = c(0, 1), evid = c(1L, 0L),
    amt = c(1, NA_real_), cmt = c("depot", "central"),
    WT = grid$WT[i], POD = grid$POD[i]
  )
}))

sim_grid <- rxode2::rxSolve(mod_typical, events = ev_grid, keep = c("WT", "POD")) |>
  as.data.frame() |>
  dplyr::group_by(id) |>
  dplyr::slice_tail(n = 1) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    cl_relerr = abs(cl / published_cl(WT, POD) - 1),
    vc_relerr = abs(vc / published_vc(WT) - 1)
  )

fig3_check <- sim_grid |>
  dplyr::filter(POD == 188) |>
  dplyr::left_join(fig3, by = "WT") |>
  dplyr::mutate(cl_per_kg = cl / WT, pct_diff = 100 * (cl_per_kg / cl_per_kg_fig - 1))

stopifnot(
  nrow(sim_grid) == nrow(grid),
  all(sim_grid$cl_relerr < 1e-10),
  all(sim_grid$vc_relerr < 1e-10),
  nrow(fig3_check) == 4L,
  # Figure reading resolution is about 0.01 L/h/kg.
  all(abs(fig3_check$pct_diff) < 1.5)
)

fig3_check |>
  dplyr::select(WT, cl_per_kg, cl_per_kg_fig, pct_diff) |>
  dplyr::rename(
    "WT (kg)" = WT,
    "CL/F per kg, model (L/h/kg)" = cl_per_kg,
    "CL/F per kg, Figure 3 (L/h/kg)" = cl_per_kg_fig,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = 3, caption = "Replicates the right-hand end (POD = 188 days) of Figure 3 of Chen 2020.")
```

```{r fig3-plot, fig.cap = "Replicates Figure 3 of Chen 2020: typical CL/F per kg against days post-transplant."}
fig3_curve <- tidyr::expand_grid(WT = fig3$WT, POD = seq(1, 190, by = 1)) |>
  dplyr::mutate(cl_per_kg = published_cl(WT, POD) / WT)
ggplot(fig3_curve, aes(POD, cl_per_kg, colour = factor(WT))) +
  geom_line(linewidth = 0.9) +
  geom_point(data = fig3_check, aes(POD, cl_per_kg), colour = "black") +
  labs(x = "POD (days)", y = "CL/F (L/h/kg)", colour = "Weight (kg)") +
  theme_minimal()
```

A 1% error in the clearance coefficient must break the tight identity:

```{r fig3-mutation}
stopifnot(all(abs(sim_grid$cl / (1.01 * published_cl(sim_grid$WT, sim_grid$POD)) - 1) > 1e-3))
```

# Scale of the variability terms

Table II prints omega and sigma without saying whether they are standard
deviations or variances, and the paper quotes no CV%. For V/F the question
is moot (0.998 as an SD or as a variance gives SDs of 0.998 and 0.999). It
matters for CL/F (SD 0.627 vs 0.792) and for the residual error (0.447 vs
0.669 proportional; 70.1 vs 8.4 ng/mL additive).

The prediction-corrected VPC (Chen 2020 Figure 2) settles it. At steady
state a trough is proportional to 1/CL, so after prediction correction the
spread is carried by the CL/F eta and the residual error. The figure's
median trough is about 170 ng/mL and the simulated 97.5th-percentile band
runs from about 330 to 700 ng/mL in the early bins and about 150 to 550
ng/mL in the last bin. The 97.5th percentile of `Y = F (1 + eps1) + eps2`
with `F = 170 exp(eta_CL)` is computed deterministically on a grid below.

```{r scale-check}
pct975 <- function(om_cl, s_prop, s_add, f_typ = 170) {
  node <- seq(-6, 6, length.out = 241)
  w <- dnorm(node) / sum(dnorm(node))
  f <- f_typ * exp(om_cl * node)
  # Y | F is normal with SD sqrt((F * s_prop)^2 + s_add^2).
  sd_y <- sqrt((f * s_prop)^2 + s_add^2)
  p_above <- function(y) sum(w * pnorm(y, f, sd_y, lower.tail = FALSE))
  uniroot(function(y) p_above(y) - 0.025, c(f_typ, 50 * f_typ))$root
}
scale_tab <- tibble::tibble(
  reading = c("SD (as encoded)", "variance"),
  om_cl = c(0.627, sqrt(0.627)),
  s_prop = c(0.447, sqrt(0.447)),
  s_add = c(70.071, sqrt(70.071))
) |>
  dplyr::rowwise() |>
  dplyr::mutate(p975 = pct975(om_cl, s_prop, s_add)) |>
  dplyr::ungroup()

stopifnot(
  # The SD reading lands at the top of the published band ...
  scale_tab$p975[1] < 720,
  # ... and the variance reading sits far above anything in Figure 2.
  scale_tab$p975[2] > 900
)
scale_tab |>
  dplyr::rename(
    "Reading" = reading, "SD of eta CL/F" = om_cl,
    "Proportional SD" = s_prop, "Additive SD (ng/mL)" = s_add,
    "97.5th percentile at 170 ng/mL" = p975
  ) |>
  knitr::kable(digits = 3, caption = "97.5th percentile implied by each reading of Table II, against the Figure 2 simulated band (about 150-700 ng/mL).")
```

The variance reading puts the 97.5th percentile near 1000 ng/mL, outside
every bin of Figure 2; the SD reading is at the top of the band. The same
authors' Wang 2019 cyclosporin paper in the same journal prints omega on the
SD scale, confirmed there by the CV%s quoted in its abstract. The model
therefore takes the printed values as SDs.

# Replicating the dose-finding simulation (Table III and Figure 4)

Chen 2020 simulated 1000 virtual children at each of 5, 10, 20 and 30 kg on
2 to 8 mg/kg/day split into two doses and reported the trough median with a
15th-85th percentile interval (Table III) and the probability of a trough
between 50 and 350 ng/mL (Figure 4). The paper states neither the POD nor
the sampling day used. Two features of Table III pin them down. First, the
troughs barely depend on body weight: with CL/F scaling as WT^0.75 and dose
as WT, a steady-state trough would rise by about 57% from 5 to 30 kg, yet
Table III shows under 1%. The troughs must be early, while concentration
is still close to accumulated dose / V/F, which scales out with weight.
Second, the median at 1 mg/kg per dose is about four doses' worth of
1 mg/kg / 93.6 L/kg. The reconstruction below therefore takes an
initial-dose scenario at POD = 1 and the trough just before the fifth dose
(48 h), without residual error. These choices are the maintainers', and
they are back-solved from the table.

The model is linear in dose, so one solve at 1 mg/kg per dose is scaled to
each regimen. The IIV integral is computed on a fixed grid of eta values
(no random cohort), passed to the typical-value model as per-subject
`lcl` / `lvc`.

```{r table3-quadrature}
# The early troughs are driven mainly by V/F, so its axis gets the finer
# grid; the probability-of-target is an indicator integral and a coarse grid
# makes it step-shaped.
node_cl <- seq(-4.5, 4.5, length.out = 41)
node_v <- seq(-5, 5, length.out = 201)
w_cl <- dnorm(node_cl) / sum(dnorm(node_cl))
w_v <- dnorm(node_v) / sum(dnorm(node_v))
eta_grid <- tidyr::expand_grid(i_cl = seq_along(node_cl), i_v = seq_along(node_v)) |>
  dplyr::mutate(w = w_cl[i_cl] * w_v[i_v])

wquantile <- function(x, w, p) {
  o <- order(x)
  cw <- cumsum(w[o]) / sum(w)
  vapply(p, function(pp) x[o][which(cw >= pp)[1]], numeric(1))
}

trough_one_weight <- function(wt) {
  params <- eta_grid |>
    dplyr::mutate(
      id = dplyr::row_number(),
      lcl = log(29.2) + 0.627 * node_cl[i_cl],
      lvc = log(6550) + 0.998 * node_v[i_v]
    )
  ev <- dplyr::bind_rows(lapply(params$id, function(i) {
    tibble::tibble(
      id = i, time = c(0, 12, 24, 36, 48), evid = c(1L, 1L, 1L, 1L, 0L),
      amt = c(rep(wt, 4), NA_real_), cmt = c(rep("depot", 4), "central"),
      WT = wt, POD = 1
    )
  }))
  # rxode2 warns that a multi-subject solve has no omega; that is intended
  # here, as the between-subject spread is supplied by the eta grid.
  sim <- suppressWarnings(rxode2::rxSolve(
    mod_typical,
    events = ev,
    params = dplyr::select(params, id, lcl, lvc)
  )) |>
    as.data.frame() |>
    dplyr::filter(time == 48)
  tibble::tibble(WT = wt, Cc = sim$Cc, w = params$w[match(sim$id, params$id)])
}

troughs <- dplyr::bind_rows(lapply(c(5, 10, 20, 30), trough_one_weight))
stopifnot(nrow(troughs) == 4L * nrow(eta_grid))

regimens <- c(2, 3, 4, 5, 6, 7, 8)
sim_tab <- tidyr::expand_grid(WT = c(5, 10, 20, 30), mgkgday = regimens) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    q = list(wquantile(
      troughs$Cc[troughs$WT == WT] * mgkgday / 2,
      troughs$w[troughs$WT == WT],
      c(0.15, 0.5, 0.85)
    )),
    q15 = q[1], med = q[2], q85 = q[3],
    pta = 100 * {
      cc <- troughs$Cc[troughs$WT == WT] * mgkgday / 2
      ww <- troughs$w[troughs$WT == WT]
      sum(ww[cc >= 50 & cc <= 350]) / sum(ww)
    }
  ) |>
  dplyr::ungroup() |>
  dplyr::select(-q)
```

```{r table3-compare}
# Chen 2020 Table III (median, 15th, 85th percentile), 5 and 30 kg columns.
# The 2 mg/kg/day 5 kg median is printed as 41.5; every other entry in the
# column is exactly proportional to dose with a 41.15 base, so 41.5 is taken
# as a transposition of 41.15.
tab3 <- tibble::tribble(
  ~WT, ~mgkgday, ~med_pub, ~q15_pub, ~q85_pub,
  5, 2, 41.15, 15.05, 104.81,
  5, 6, 123.45, 45.15, 314.43,
  30, 2, 41.54, 15.11, 108.09,
  30, 6, 124.62, 45.32, 324.29
)
# Chen 2020 Figure 4, read by the maintainers (probability, %, of a trough
# in 50-350 ng/mL); digitisation resolution about 0.3 percentage points.
fig4 <- tibble::tribble(
  ~mgkgday, ~WT5, ~WT10, ~WT20, ~WT30,
  2, 41.8, 42.0, 42.2, 42.3,
  3, 55.8, 56.1, 56.2, 56.1,
  4, 64.0, 63.6, 63.3, 62.8,
  5, 68.9, 68.4, 68.0, 67.9,
  6, 69.7, 68.8, 68.9, 68.9,
  7, 70.1, 69.7, 69.2, 69.1,
  8, 68.6, 67.9, 67.4, 67.0
) |>
  tidyr::pivot_longer(-mgkgday, names_to = "WT", values_to = "pta_pub") |>
  dplyr::mutate(WT = as.numeric(sub("WT", "", WT)))

tab3_check <- dplyr::inner_join(sim_tab, tab3, by = c("WT", "mgkgday")) |>
  dplyr::mutate(
    med_diff = 100 * (med / med_pub - 1),
    q15_diff = 100 * (q15 / q15_pub - 1),
    q85_diff = 100 * (q85 / q85_pub - 1)
  )
fig4_check <- dplyr::inner_join(sim_tab, fig4, by = c("WT", "mgkgday")) |>
  dplyr::mutate(pta_diff = pta - pta_pub)

stopifnot(
  nrow(tab3_check) == 4L,
  nrow(fig4_check) == 28L,
  # The published percentiles come from 1000 simulated children, so they
  # carry about 5% Monte Carlo error (15th/85th) and 3% (median).
  all(abs(tab3_check$med_diff) < 6),
  all(abs(tab3_check$q15_diff) < 10),
  all(abs(tab3_check$q85_diff) < 12),
  # Figure 4: about 1.5 points of Monte Carlo error on each published value.
  abs(median(fig4_check$pta_diff)) < 2,
  max(abs(fig4_check$pta_diff)) < 5,
  # Weight invariance, the feature that fixes the scenario.
  max(sim_tab$med[sim_tab$mgkgday == 2]) / min(sim_tab$med[sim_tab$mgkgday == 2]) < 1.02
)

tab3_check |>
  dplyr::select(WT, mgkgday, med, med_pub, q15, q15_pub, q85, q85_pub) |>
  dplyr::rename(
    "WT (kg)" = WT, "Dose (mg/kg/day)" = mgkgday,
    "Median, model" = med, "Median, Table III" = med_pub,
    "15th, model" = q15, "15th, Table III" = q15_pub,
    "85th, model" = q85, "85th, Table III" = q85_pub
  ) |>
  knitr::kable(digits = 2, caption = "Replicates Table III of Chen 2020 (trough ng/mL, 48 h after the first dose at POD = 1).")
```

```{r fig4-plot, fig.cap = "Replicates Figure 4 of Chen 2020: probability of a trough in 50-350 ng/mL. Lines, this model; points, values read from Figure 4."}
ggplot(fig4_check, aes(WT, pta, colour = factor(mgkgday))) +
  geom_line() +
  geom_point(aes(y = pta_pub), shape = 1, size = 2) +
  labs(x = "Weight (kg)", y = "Probability of target attainment (%)", colour = "mg/kg/day") +
  theme_minimal()
```

The model reproduces the paper's conclusion: 6 and 7 mg/kg/day give the
highest probability of a 50-350 ng/mL trough, and at 7 mg/kg/day the 85th
percentile exceeds 350 ng/mL at every weight
(`r paste(round(sim_tab$q85[sim_tab$mgkgday == 7]), collapse = ", ")` ng/mL
for 5, 10, 20 and 30 kg), which is why 6 mg/kg/day was chosen.

# Steady-state PKNCA check

At a fixed POD the model is linear and time-invariant, so at steady state
`AUC_tau = Dose / (CL/F)` exactly. The typical 7.6 kg child (cohort median)
at POD 51.5 days receives 6 mg/kg/day as 22.8 mg every 12 h for 60 days;
PKNCA is run on the final dosing interval for each of the four simulated
weights.

```{r pknca-ss}
wts <- c(5, 7.6, 20, 30)
tau <- 12
n_dose <- 120
t_last <- (n_dose - 1) * tau
obs_t <- t_last + seq(0, tau, by = 0.25)
ev_ss <- dplyr::bind_rows(lapply(seq_along(wts), function(i) {
  dplyr::bind_rows(
    tibble::tibble(id = i, time = (0:(n_dose - 1)) * tau, evid = 1L,
                   amt = 3 * wts[i], cmt = "depot"),
    tibble::tibble(id = i, time = obs_t, evid = 0L, amt = NA_real_, cmt = "central")
  ) |>
    dplyr::mutate(WT = wts[i], POD = 51.5, treatment = paste0(wts[i], " kg"))
})) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim_ss <- rxode2::rxSolve(mod_typical, events = ev_ss, keep = c("WT", "treatment")) |>
  as.data.frame()

conc_ss <- sim_ss |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(time = time - t_last) |>
  dplyr::select(id, treatment, time, Cc)
dose_ss <- ev_ss |>
  dplyr::filter(evid == 1, time == t_last) |>
  dplyr::mutate(time = 0) |>
  dplyr::select(id, treatment, time, amt)

nca_ss <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_ss, Cc ~ time | treatment + id, concu = "ng/mL", timeu = "h"),
  PKNCA::PKNCAdose(dose_ss, amt ~ time | treatment + id, doseu = "mg"),
  intervals = data.frame(start = 0, end = tau, auclast = TRUE, cmax = TRUE, cmin = TRUE)
))
nca_ss_wide <- as.data.frame(nca_ss$result) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::mutate(
    WT = as.numeric(sub(" kg", "", treatment)),
    auc_expected = 3 * WT / published_cl(WT, 51.5) * 1000,
    pct_diff = 100 * (auclast / auc_expected - 1)
  )

# Deterministic solve against its own closed form: a tight bound is correct.
stopifnot(nrow(nca_ss_wide) == 4L, all(abs(nca_ss_wide$pct_diff) < 0.5))

nca_ss_wide |>
  dplyr::arrange(WT) |>
  dplyr::select(treatment, cmax, cmin, auclast, auc_expected, pct_diff) |>
  dplyr::rename(
    "Weight" = treatment, "Cmax (ng/mL)" = cmax, "Cmin (ng/mL)" = cmin,
    "AUC0-12 PKNCA (ng*h/mL)" = auclast, "Dose/(CL/F) (ng*h/mL)" = auc_expected,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = 2, caption = "Typical-value steady state at 6 mg/kg/day and POD 51.5 days.")
```

Chen 2020 reports no NCA parameters, so there is no published NCA table to
compare against.

# Stochastic trough profile with time-varying POD

A cohort of 200 children at the median weight (7.6 kg) starts 6 mg/kg/day
on POD 1, with POD advancing with the simulation clock, and troughs are
recorded daily for 60 days (residual error included).

```{r vpc, fig.cap = "Simulated troughs (median and 15th-85th percentile) at 6 mg/kg/day from POD 1; the shaded band is the 50-350 ng/mL target."}
rxode2::rxSetSeed(20200601)
n_sub <- 200
days <- 60
dose_t <- seq(0, days * 24 - tau, by = tau)
obs_d <- seq(24, days * 24, by = 24) - 1e-6
ev_vpc <- dplyr::bind_rows(
  tidyr::expand_grid(id = seq_len(n_sub), time = dose_t) |>
    dplyr::mutate(evid = 1L, amt = 3 * 7.6, cmt = "depot"),
  tidyr::expand_grid(id = seq_len(n_sub), time = obs_d) |>
    dplyr::mutate(evid = 0L, amt = NA_real_, cmt = "central")
) |>
  dplyr::mutate(WT = 7.6, POD = 1 + time / 24) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim_vpc <- rxode2::rxSolve(mod, events = ev_vpc, keep = "POD") |>
  as.data.frame()
vpc_sum <- sim_vpc |>
  dplyr::filter(!is.na(sim)) |>
  dplyr::mutate(day = round(time / 24)) |>
  dplyr::group_by(day) |>
  dplyr::summarise(
    q15 = quantile(sim, 0.15), med = median(sim), q85 = quantile(sim, 0.85),
    .groups = "drop"
  )
ggplot(vpc_sum, aes(day, med)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 50, ymax = 350, alpha = 0.1, fill = "darkgreen") +
  geom_ribbon(aes(ymin = q15, ymax = q85), alpha = 0.3) +
  geom_line(linewidth = 0.9) +
  labs(x = "Day of therapy", y = "Ciclosporin trough (ng/mL)") +
  theme_minimal()
```

Because clearance is near zero immediately after transplant and rises with
POD, troughs accumulate over the first weeks and then fall as clearance
catches up. The paper's dose recommendation concerns the initial dose only;
it was not designed for, and should not be read as, a maintenance regimen.

# Assumptions and deviations

* **Omega and sigma scale.** Table II does not name its scale. The values
  are read as standard deviations, on the evidence of the pcVPC (above) and
  the same group's convention in Wang 2019. For V/F the two readings are
  indistinguishable.
* **Table III / Figure 4 scenario.** The paper does not state the POD or
  sampling time of its dose-finding simulation. POD = 1 with the trough 48 h
  after the first of twice-daily doses, without residual error, was
  back-solved from the weight invariance and dose scaling of Table III, and
  it reproduces Figure 4. No model parameter was adjusted.
* **Table III typo.** The 5 kg, 2 mg/kg/day median is printed as 41.5 ng/mL;
  the rest of the column is exactly dose-proportional to 41.15, which is used
  here.
* **POD at zero.** The power form makes CL/F zero at POD = 0. The observed
  data start at POD = 1; supply POD >= 1.
* **Covariates without canonical columns.** Mycophenolate mofetil and
  tacrolimus co-medication, mean corpuscular hemoglobin and mean corpuscular
  hemoglobin concentration were screened but have no canonical column; they
  are recorded in the model's `population` notes only.
* **Figure readings.** Figure 3 and Figure 4 values were read by the
  maintainers from the published figures.
