---
title: "Sirolimus (Cheng 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sirolimus (Cheng 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Cheng_2020_sirolimus"))$reference`
* Description: `r rxode2::rxode(readModelDb("Cheng_2020_sirolimus"))$description`
* Article: <https://doi.org/10.1177/2058738420934936> (open access, PMC7388097)

Cheng and colleagues fit a one-compartment model with first-order absorption to
107 routine therapeutic-drug-monitoring (TDM) trough concentrations from 27
Chinese children with refractory immune cytopenia who took oral sirolimus. The
analysis used Phoenix NLME 1.3. Of 19 screened covariates, body weight and total
bilirubin were retained, both on apparent clearance. The paper's practical
output is Table 4, a table of initial daily doses by body weight and total
bilirubin.

# Population

The single-centre pilot study at Beijing Children's Hospital ran from January
2016 to December 2017 (Cheng 2020 Table 1). The 27 children (18 male, 9 female)
were 8.16 +/- 3.60 years old (range 1-15) and weighed 27.03 +/- 10.87 kg (range
7-43; median 28.50 kg). Mean total bilirubin was 12.13 +/- 7.35 umol/L (median
11.29 umol/L). Sirolimus gelatin capsules were given orally once daily. The
starting dose was 1.5 mg/m^2, and later doses were titrated by TDM to a trough
of 5-15 ng/mL. Every concentration is a steady-state trough drawn 0.5 h before
the next dose, at least 7 days after the first dose. Whole blood was assayed by
fluorescence polarization immunoassay (Abbott TDx/FLx).

The same information is available programmatically via
`readModelDb("Cheng_2020_sirolimus")()$population`.

# Source trace

| Element | Value | Source |
|---|---|---|
| Structure: one compartment, first-order absorption and elimination | -- | Methods, 'PPK analysis'; Results |
| `lcl` (CL/F at 28.5 kg, TBIL 11.29 umol/L) | 5.63 L/h | Table 3; final-model equation (Results) |
| `lvc` (V/F) | 144.16 L | Table 3 |
| `lka` (Ka, held constant) | 0.7521 1/h | Methods, 'PPK analysis' (Table 3 prints 0.75, RSE 0) |
| `e_tbili_cl` | -0.32 | Table 3 'f CL-TBIL'; final-model equation |
| `e_wt_cl` | 0.50 | Table 3 'f CL-WT'; final-model equation |
| Reference total bilirubin | 11.29 umol/L | Results: '11.29 is median TBIL' |
| Reference body weight | 28.50 kg | Results: 'The median weight of children was 28.50 kg' |
| `etalcl` | 0.0353 (log-scale variance) | Abstract: IIV of CL/F 3.53% (see below) |
| `etalvc` | 0.0727 (log-scale variance) | Abstract: IIV of V/F 7.27% (see below) |
| `propSd` | 0.2245 | Abstract: proportional error 22.45%; Table 3 Sigma 0.22 |
| No covariate on V/F | -- | Results: 'V was by no co-variants influenced' |

The final-model equation printed in Results is

  CL (L/h) = 5.63 x (TBIL / 11.29)^-0.32 x (WT / 28.50)^0.5 x exp(eta_CL).

## Scale of the inter-individual variability

Table 3 has no omega rows. The only report of the between-subject variability is
the abstract: "Inter-individual variabilities for CL/F and V/F were 3.53% and
7.27%, respectively. The intra-individual variability of proportional error
model was 22.45%." Two readings are possible:

* **Omega diagonal times 100 (used).** Phoenix NLME prints the Omega matrix as
  log-scale variances and the residual term as a standard deviation (`stdev0`).
  Table 3 gives the residual term as 0.22, and the abstract turns it into
  "22.45%" by multiplying the printed estimate by 100. Applying the same
  convention to the omegas gives omega^2 = 0.0353 and 0.0727, which are apparent
  CVs of about 19% and 27%.
* **Literal CVs.** omega^2 = log(1 + CV^2) = 0.00125 and 0.00527, which are
  apparent CVs of 3.5% and 7.3%.

No other result in the paper separates the two readings: there is no RSE, no
shrinkage and no published prediction interval. The model uses the first
reading because it treats the omegas and the residual term the same way. This
only affects the stochastic spread in the simulations below. The typical-value
checks do not depend on it.

```{r omega-scale}
omega_pct <- c(CL = 3.53, V = 7.27)
data.frame(
  parameter = names(omega_pct),
  omega2_used = omega_pct / 100,
  apparent_cv_used_pct = 100 * sqrt(exp(omega_pct / 100) - 1),
  omega2_literal_cv = log(1 + (omega_pct / 100)^2)
) |>
  knitr::kable(digits = 5)
```

# Virtual cohort

Table 4 splits patients into four body-weight bands (5-15, 15-25, 25-35 and
35-45 kg) and two total-bilirubin bands (3.42-20.50 and > 20.50 umol/L). It
gives a daily initial dose for each of the eight cells. Each cell below has 200
virtual children, with body weight drawn uniformly within the band. Total
bilirubin is drawn uniformly over 3.42-20.50 umol/L in the normal band and over
20.50-40 umol/L in the high band. The paper gives no upper bound for the high
band, so 40 umol/L is an assumption.

```{r cohort}
mod <- readModelDb("Cheng_2020_sirolimus")
ui <- rxode2::rxode(mod)
theta <- ui$theta

tab4 <- tibble::tribble(
  ~wt_lo, ~wt_hi, ~tbili_band, ~dose_mg,
  5, 15, "normal", 0.75,
  15, 25, "normal", 1.5,
  25, 35, "normal", 1.75,
  35, 45, "normal", 2.0,
  5, 15, "high", 0.5,
  15, 25, "high", 0.75,
  25, 35, "high", 1.0,
  35, 45, "high", 1.5
) |>
  mutate(
    cell = sprintf("WT %g-%g kg, TBIL %s", wt_lo, wt_hi, tbili_band),
    cell_id = row_number()
  )

n_per_cell <- 200
tau <- 24
n_dose <- 56
t_end <- n_dose * tau

rxode2::rxSetSeed(20200701)
subj <- tab4[rep(seq_len(nrow(tab4)), each = n_per_cell), ] |>
  mutate(
    id = row_number(),
    WT = runif(n(), wt_lo, wt_hi),
    TBILI = ifelse(tbili_band == "normal", runif(n(), 3.42, 20.5), runif(n(), 20.5, 40))
  )
stopifnot(!anyDuplicated(subj$id))
```

# Simulation

Each child takes the Table 4 dose once daily for 56 days (1344 h). The typical
half-life is `r signif(log(2) / (exp(theta[["lcl"]]) / exp(theta[["lvc"]])), 3)`
h. In the lightest, high-bilirubin children it is three to four times longer, and it is longer still for those who also clear slowly by chance, which is why the run is eight weeks.
Observations are taken 0.5 h before every dose, which is the paper's sampling
time, and densely over the last dosing interval for NCA.

```{r simulate}
obs_times <- sort(unique(c(
  0,
  seq(tau, t_end, by = tau) - 0.5,
  seq(t_end - tau, t_end, by = 0.5)
)))

make_events <- function(s) {
  dose <- data.frame(
    id = s$id, time = seq(0, t_end - tau, by = tau), amt = s$dose_mg,
    evid = 1L, cmt = "depot"
  )
  obs <- data.frame(id = s$id, time = obs_times, amt = 0, evid = 0L, cmt = "central")
  ev <- rbind(dose, obs)
  ev$WT <- s$WT
  ev$TBILI <- s$TBILI
  ev$cell_id <- s$cell_id
  ev
}
events <- do.call(rbind, lapply(split(subj, subj$id), make_events))
events <- events[order(events$id, events$time, -events$evid), ]

sim <- rxode2::rxSolve(ui, events, keep = "cell_id", returnType = "data.frame")
stopifnot(length(unique(sim$id)) == nrow(subj))
sim <- sim |> left_join(tab4 |> select(cell_id, cell, tbili_band, dose_mg), by = "cell_id")
```

# Replicate published figures

## Figure 3 - trough concentrations over time

Figure 3 of Cheng 2020 is a VPC of the observed troughs. The observed data are
not published, so the figure below shows the simulated trough percentiles over
the eight simulated weeks in each cell of Table 4, with the 5-15 ng/mL target band.

```{r fig-troughs, fig.height = 6}
troughs <- sim |>
  filter(time > 0, (time + 0.5) %% tau == 0)

troughs |>
  group_by(cell, time) |>
  summarise(
    p05 = quantile(Cc, 0.05), p50 = median(Cc), p95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time / 24, p50)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 5, ymax = 15, alpha = 0.15, fill = "darkgreen") +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.3) +
  geom_line() +
  facet_wrap(~cell, ncol = 4) +
  labs(
    x = "Day", y = "Trough sirolimus (ng/mL)",
    caption = "Simulated 5th/50th/95th percentiles; shaded band is the 5-15 ng/mL target (cf. Figure 3 of Cheng 2020)."
  )
```

## Table 4 - initial dose by body weight and total bilirubin

The paper does not report attainment probabilities for Table 4, so the table is
checked two ways. First, a typical child at the midpoint of each weight band
must have a steady-state trough inside 5-15 ng/mL on the recommended dose. The
typical child has median total bilirubin (11.29 umol/L) in the normal band and
25 umol/L in the high band. Second, the table reports the simulated fraction of
children whose steady-state trough falls in the target window.

```{r table4}
# Closed-form steady-state trough of a one-compartment oral model, sampled
# 0.5 h before the next dose.
css_oral <- function(dose, cl, v, ka, tau, t) {
  k <- cl / v
  1000 * dose * ka / (v * (ka - k)) *
    (exp(-k * t) / (1 - exp(-k * tau)) - exp(-ka * t) / (1 - exp(-ka * tau)))
}
typ <- tab4 |>
  mutate(
    WT = (wt_lo + wt_hi) / 2,
    TBILI = ifelse(tbili_band == "normal", 11.29, 25),
    cl = exp(theta[["lcl"]]) * (TBILI / 11.29)^theta[["e_tbili_cl"]] * (WT / 28.5)^theta[["e_wt_cl"]],
    v = exp(theta[["lvc"]]),
    trough_typ = css_oral(dose_mg, cl, v, exp(theta[["lka"]]), tau, tau - 0.5)
  )

ss <- troughs |>
  filter(time == t_end - 0.5) |>
  group_by(cell) |>
  summarise(
    median_trough = median(Cc),
    pct_in_5_15 = 100 * mean(Cc >= 5 & Cc <= 15),
    pct_below_5 = 100 * mean(Cc < 5),
    pct_above_15 = 100 * mean(Cc > 15),
    .groups = "drop"
  )

typ |>
  select(cell, dose_mg, trough_typ) |>
  left_join(ss, by = "cell") |>
  dplyr::rename(
    "Table 4 cell" = cell, "Daily dose (mg)" = dose_mg,
    "Typical trough (ng/mL)" = trough_typ, "Median simulated trough" = median_trough,
    "% in 5-15" = pct_in_5_15, "% < 5" = pct_below_5, "% > 15" = pct_above_15
  ) |>
  knitr::kable(digits = c(0, 2, 1, 1, 1, 1, 1))

# Deterministic typical-value check: every recommended dose puts the typical
# child of its cell inside the target window.
stopifnot(all(typ$trough_typ > 5 & typ$trough_typ < 15))
```

Every Table 4 dose puts the typical child of its cell inside the target window.
The median simulated trough in each cell is close to the typical value, and most
simulated children in each cell are in the window.

## Closed-form check of the ODE solution

The ODE solution at steady state must match the closed-form trough for the same
typical parameters. Both sides use the same parameters, so the only difference
is numerical error, and the bound can be tight.

```{r closed-form}
typ$id <- seq_len(nrow(typ))
ev_typ <- do.call(rbind, lapply(split(typ, typ$id), make_events))
ev_typ <- ev_typ[order(ev_typ$id, ev_typ$time, -ev_typ$evid), ]
sim_typ <- rxode2::rxSolve(rxode2::zeroRe(ui), ev_typ, returnType = "data.frame", omega = NA) |>
  filter(time == t_end - 0.5) |>
  arrange(id)
chk <- data.frame(ode = sim_typ$Cc, closed = typ$trough_typ)
chk$pct_diff <- 100 * (chk$ode - chk$closed) / chk$closed
knitr::kable(cbind(cell = typ$cell, round(chk, 4)))
# 56 days is > 30 typical half-lives even in the slowest cell, so the ODE
# trough has converged to steady state.
stopifnot(all(abs(chk$pct_diff) < 0.5))
```

# PKNCA validation

PKNCA is run on the last dosing interval (1320-1344 h) of every simulated child,
grouped by Table 4 cell. At steady state, AUC over one dosing interval equals
Dose / CL for each individual, with the dose in mg and CL in L/h (times 1000 for
ng.h/mL). This gives a per-subject identity. The only error is the trapezoidal
error on the 0.5 h grid.

```{r pknca}
conc <- sim |>
  filter(time >= t_end - tau, !is.na(Cc)) |>
  select(id, time, Cc, cell)
dose_df <- events |>
  filter(evid == 1, time == t_end - tau) |>
  select(id, time, amt) |>
  left_join(subj |> select(id, cell), by = "id")

conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ time | cell + id, concu = "ng/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | cell + id, doseu = "mg")
intervals <- data.frame(start = t_end - tau, end = t_end, auclast = TRUE, cmax = TRUE, cmin = TRUE, tmax = TRUE)
nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_res <- as.data.frame(nca$result)

nca_res |>
  group_by(cell, PPTESTCD) |>
  summarise(median = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  dplyr::rename(
    "Table 4 cell" = cell, "AUCtau (ng.h/mL)" = auclast, "Cmax (ng/mL)" = cmax,
    "Cmin (ng/mL)" = cmin, "Tmax (h)" = tmax
  ) |>
  knitr::kable(digits = 2)

# Per-subject identity AUCtau = 1000 * Dose / CL_i.
cl_i <- sim |>
  filter(time == t_end) |>
  select(id, cl)
auc_chk <- nca_res |>
  filter(PPTESTCD == "auclast") |>
  left_join(cl_i, by = "id") |>
  left_join(subj |> select(id, dose_mg), by = "id") |>
  mutate(expected = 1000 * dose_mg / cl, pct_diff = 100 * (PPORRES - expected) / expected)
stopifnot(nrow(auc_chk) == nrow(subj))
# Same drawn parameters on both sides; residual difference is trapezoidal error,
# plus a tiny carry-over for the slowest-clearing children.
stopifnot(abs(median(auc_chk$pct_diff)) < 0.5, quantile(abs(auc_chk$pct_diff), 0.9) < 1)
summary(auc_chk$pct_diff)
```

Cheng 2020 reports no NCA parameters (all the data are troughs), so there is no
published NCA to compare against. The PKNCA results are checked against the
exact steady-state identity above.

# Assumptions and deviations

* **IIV scale.** The abstract's 3.53% and 7.27% are read as Phoenix omega^2 x
  100 (log-scale variances 0.0353 and 0.0727). This is the same x100
  convention the abstract uses for the residual term (Table 3 Sigma 0.22,
  written as "22.45%"). The literal-CV alternative is shown in the "Scale of the
  inter-individual variability" section. No result in the paper discriminates
  between them.
* **Residual error.** `propSd = 0.2245` uses the abstract's more precise value.
  Table 3 rounds it to 0.22.
* **Ka.** Ka is 0.7521 1/h (Methods) rather than the 0.75 printed in Table 3.
  The Methods value is the one that was held constant, and the Discussion quotes
  it as 0.752.
* **Dosing interval.** The Methods say sirolimus "was given orally daily", so
  every simulation uses once-daily dosing. Table 4 doses are read as daily
  doses.
* **Table 4 covariate distributions.** Uniform draws within each band. The
  high-bilirubin band has no printed upper bound, and 40 umol/L is assumed. The
  typical-value check uses band-midpoint weights and 11.29 / 25 umol/L total
  bilirubin.
* **Assay range.** The Methods give the assay's lower limit of quantification as
  25 ng/mL, which is above the whole 5-15 ng/mL target range the troughs were
  titrated to. This is probably a typographical error (FPIA sirolimus assays
  usually quantify down to about 2.5 ng/mL). It has no effect on the model,
  which does not model censoring.
* **Figure 3 (VPC)** cannot be reproduced without the observed data. A
  simulated trough-percentile figure is shown instead.
