---
title: "TB combination therapy MTP-GPDI in BALB/c mice (Chen 2017)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TB combination therapy MTP-GPDI in BALB/c mice (Chen 2017)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Chen_2017_TB_MTP_GPDI_mouse"))$reference`
* Description: `r rxode2::rxode(readModelDb("Chen_2017_TB_MTP_GPDI_mouse"))$description`
* Article: <https://doi.org/10.1002/psp4.12226>

# Population

The model was developed from female BALB/c mice (13-15 weeks old, 20-25 g)
intratracheally infected with the M. tuberculosis Beijing VN 2002-1585 genotype
strain (~9.5e4 CFU inoculum) at the Erasmus MC (Rotterdam, NL). Forty-nine
TB-infected mice contributed sparse PK (one plasma sample per animal at 1, 4,
or 8 h post-dose); an additional eighteen healthy mice contributed a rifampicin
PK sub-study at 10 and 160 mg/kg. CFU sampling lasted 1-4 weeks for the
rifampicin and isoniazid monotherapy arms, 1 week for ethambutol and
pyrazinamide (the animals did not survive longer), and 1-24 weeks for the
combination arms (R10H25, R10H25Z150, R10H25Z150E100). All drugs were given by
once-daily oral gavage on a 5-day-on / 2-day-off weekly schedule.

The `population` metadata field of the model carries the same information
programmatically:

```{r population}
rxode2::rxode(readModelDb("Chen_2017_TB_MTP_GPDI_mouse"))$population
```

# Source trace

Every `ini()` entry carries an in-file comment pointing to the Chen 2017 (or
upstream) source location. The table below collects them for reviewer use.

| Block | Parameter | Value | Units | Source |
|-------|-----------|-------|-------|--------|
| RIF PK | `lka_rif` | log(6.23) | 1/h | Chen 2017 Table 1 (TB-infected, R5/R10/R20) |
| RIF PK | `lcl_rif` | log(234) | mL/h/kg | Chen 2017 Table 1 |
| RIF PK | `lvc_rif` | log(706) | mL/kg | Chen 2017 Table 1 |
| INH PK | `lka_inh` | log(12.6) FIX | 1/h | Chen 2017 Table 1 footnote a (fixed from Chen 2016 EJPS) |
| INH PK | `lcl_inh_lowdose` | log(612) | mL/h/kg | Chen 2017 Table 1 (lowest dose H12.5) |
| INH PK | `slope_inh` | 7.50e-3 | 1/(mg/kg) | Chen 2017 Table 1 (dose-dependence slope) |
| INH PK | `lvc_inh` | log(811) | mL/kg | Chen 2017 Table 1 |
| EMB PK | `lka_emb` | log(0.87) FIX | 1/h | Chen 2017 Table 1 footnote a |
| EMB PK | `lcl_emb` | log(3400) | mL/h/kg | Chen 2017 Table 1 |
| EMB PK | `lvc_emb` | log(1500) | mL/kg | Chen 2017 Table 1 |
| EMB PK | `lq_emb` | log(2530) | mL/h/kg | Chen 2017 Table 1 |
| EMB PK | `lvp_emb` | log(4690) | mL/kg | Chen 2017 Table 1 |
| PZA PK | `lka_pza` | log(2.84) FIX | 1/h | Chen 2017 Table 1 footnote a |
| PZA PK | `lcl_pza` | log(273.71) | mL/h/kg | Chen 2017 Table 1 |
| PZA PK | `lvc_pza` | log(525.1) | mL/kg | Chen 2017 Table 1 |
| MTP | `f0_init` | log(20100) | log(CFU/lungs) | Chen 2017 Table 2 |
| MTP | `s0_init` | log(119000) | log(CFU/lungs) | Chen 2017 Table 2 |
| MTP | `lkg` | log(0.034) | 1/h | Chen 2017 Table 2 |
| MTP | `lkfs_lin` | log(6.65e-5) | 1/h^2 | Chen 2017 Table 2 |
| MTP | `lksf` | log(6.03e-4) FIX | 1/h | Chen 2017 Table 2 footnote a (fixed from Clewe 2016) |
| MTP | `lkfn` | log(3.74e-8) FIX | 1/h | Chen 2017 Table 2 footnote a |
| MTP | `lksn` | log(7.73e-3) FIX | 1/h | Chen 2017 Table 2 footnote a |
| MTP | `lkns` | log(5.11e-5) FIX | 1/h | Chen 2017 Table 2 footnote a |
| RIF drug effects | `lkg_r` | log(0.0022) | mL/(h*ug) | Chen 2017 Table 2 (linear F-growth inhibition) |
| RIF drug effects | `loo_f_r` | log(0.019) | 1/h | Chen 2017 Table 2 (on/off F-death) |
| RIF drug effects | `lks_r` | log(0.0137) | mL/(h*ug) | Chen 2017 Table 2 (linear S-death) |
| RIF drug effects | `lkn_r` | log(0.0033) | mL/(h*ug) | Chen 2017 Table 2 (linear N-death) |
| INH drug effects | `lkf_h` | log(0.055) | mL/(h*ug) | Chen 2017 Table 2 (linear F-death) |
| INH drug effects | `lks_h` | log(0.00047) | mL/(h*ug) | Chen 2017 Table 2 (linear S-death) |
| INH drug effects | `lkn_h` | log(0.0012) | mL/(h*ug) | Chen 2017 Table 2 (linear N-death) |
| GPDI | `int_s_rh` | 4.49 | (fractional) | Chen 2017 Table 2 (RIF-INH on S death) |
| GPDI | `int_n_rh` | 0.32 | (fractional) | Chen 2017 Table 2 (RIF-INH on N death) |
| GPDI | `int_n_re` | -0.15 | (fractional) | Chen 2017 Table 2 (RIF-EMB on N death) |
| Residual | `addSd` | sqrt(1.23)*log(10) FIX | log(CFU/lungs) | Chen 2017 Table 2 (sigma^2 = 1.23 on log10 scale) |

Equation references:

* MTP three-state ODEs for fast (F), slow (S), non-multiplying (N): Chen 2017
  Methods, Eq. 1-3 for natural growth and Eq. 10-12 for the final MTP-GPDI
  model.
* GPDI on/off (joint INT_AB = INT_BA, EC50_INT -> 0) reduction with linear
  monotherapy slope: Chen 2017 Methods, Eq. 9.
* Isoniazid dose-dependent clearance CL_inh = CL_inh_lowest * (1 - slope_sigma *
  (Dose - Dose_lowest)): Chen 2017 Table 1 footnote b.
* MTP natural-growth transfer rates kSF, kFN, kSN, kNS fixed to the in vitro
  Clewe 2016 estimates: Chen 2017 Methods (Evaluation of drug effects in
  monotherapy section) and Table 2 footnote a.
* Absorption rate constants of INH/EMB/PZA fixed to the mouse-population values
  of Chen 2016 (Eur J Pharm Sci 93:319-333): Chen 2017 Table 1 footnote a.

# Virtual cohort

The simulations below reproduce the typical-value trajectories of Chen 2017
Figure 2 (no IIV) and demonstrate the qualitative behaviour of natural growth,
RIF monotherapy, INH monotherapy, the RIF+INH antagonistic combination
(R10H25), and the four-drug R10H25Z150E100 combination that involves the
RIF-EMB synergistic interaction on N-state bacteria.

```{r cohort}
mouse_kg <- 0.022  # 22 g body weight, mid-range of Chen 2017 Methods (20-25 g)

# Build a per-regimen dosing schedule. Each regimen runs 5 days on / 2 days off
# for the chosen number of weeks; doses are oral (cmt = depot_<drug>). Per
# Chen 2017 Figure 2, monotherapy CFU is reported up to day 28; combination
# therapy up to day 168 (24 weeks). We simulate 28 days of monotherapy and 28
# days of combination therapy so the figure matches Chen 2017 Figure 2.

oral_doses <- function(start_day, n_weeks, drug_cmt, dose_mg_per_kg) {
  if (dose_mg_per_kg == 0) return(NULL)
  # 5 days on, 2 days off, for n_weeks
  days_on <- as.vector(outer(0:(n_weeks - 1), 0:4, function(w, d) 7 * w + d))
  tibble(
    time = (start_day + days_on) * 24,             # hours
    amt  = dose_mg_per_kg * mouse_kg * 1000,       # mg/kg -> ug (mass dose into ug/mL conc system)
    evid = 1L,
    cmt  = drug_cmt
  )
}

make_regimen <- function(id, regimen, doses_rif, doses_inh, doses_emb, doses_pza,
                         n_weeks = 4L, obs_day_grid = c(0, 1, 2, 3, 7, 14, 21, 28)) {
  events <- bind_rows(
    oral_doses(0L, n_weeks, "depot_rif", doses_rif),
    oral_doses(0L, n_weeks, "depot_inh", doses_inh),
    oral_doses(0L, n_weeks, "depot_emb", doses_emb),
    oral_doses(0L, n_weeks, "depot_pza", doses_pza)
  )
  obs <- tibble(
    time = obs_day_grid * 24,
    amt  = 0,
    evid = 0L,
    cmt  = "Cc"
  )
  bind_rows(events, obs) %>%
    mutate(
      id            = id,
      regimen       = regimen,
      DOSE_INH_MGKG = doses_inh
    ) %>%
    arrange(time, desc(evid))
}

events <- bind_rows(
  make_regimen(1L, "Natural growth",
               doses_rif = 0,  doses_inh = 0,  doses_emb = 0,   doses_pza = 0,
               obs_day_grid = c(0, 1, 3, 7, 14, 21, 28)),
  make_regimen(2L, "R10 (RIF 10 mg/kg)",
               doses_rif = 10, doses_inh = 0,  doses_emb = 0,   doses_pza = 0),
  make_regimen(3L, "H25 (INH 25 mg/kg)",
               doses_rif = 0,  doses_inh = 25, doses_emb = 0,   doses_pza = 0),
  make_regimen(4L, "R10H25",
               doses_rif = 10, doses_inh = 25, doses_emb = 0,   doses_pza = 0),
  make_regimen(5L, "R10H25Z150",
               doses_rif = 10, doses_inh = 25, doses_emb = 0,   doses_pza = 150),
  make_regimen(6L, "R10H25Z150E100",
               doses_rif = 10, doses_inh = 25, doses_emb = 100, doses_pza = 150)
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod <- readModelDb("Chen_2017_TB_MTP_GPDI_mouse")
sim <- rxode2::rxSolve(
  mod, events = events,
  keep = c("regimen", "DOSE_INH_MGKG"),
  returnType = "data.frame"
) %>% as_tibble()
```

# Replicate published figures

```{r figure-2}
# Replicates Figure 2 of Chen 2017: predicted log10(CFU/lungs) by regimen for
# the first 4 weeks of treatment. The natural-growth cohort is the no-drug
# baseline; R10 and H25 are monotherapy; R10H25 shows the RIF-INH antagonistic
# interaction on S and N states; R10H25Z150E100 adds the RIF-EMB synergistic
# interaction on N state.

sim_obs <- sim %>%
  mutate(
    day = time / 24,
    log10_total = log10(pmax(fbugs + sbugs + nbugs, 1))
  )

ggplot(sim_obs, aes(day, log10_total, colour = regimen)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  scale_y_continuous(name = "log10(CFU/lungs)") +
  scale_x_continuous(name = "Days post-treatment start",
                     breaks = c(0, 7, 14, 21, 28)) +
  labs(
    title = "Chen 2017 Figure 2 -- typical-value MTP-GPDI predictions",
    caption = "RIF-INH antagonism reduces the slope of S/N kill in R10H25; RIF-EMB synergy enhances N-state kill in R10H25Z150E100."
  ) +
  theme_minimal()
```

# Drug plasma profiles

The plasma concentration trajectories of the four drugs at the canonical
combination dose levels (R10H25Z150E100) are shown below. Each panel is
generated from the typical-value PK of Chen 2017 Table 1, which is the
disposition that feeds the MTP-GPDI bacterial killing.

```{r figure-1-pk}
sim_pk <- sim %>%
  filter(regimen == "R10H25Z150E100") %>%
  mutate(day = time / 24) %>%
  select(day, Cc_rif, Cc_inh, Cc_emb, Cc_pza) %>%
  pivot_longer(c(Cc_rif, Cc_inh, Cc_emb, Cc_pza),
               names_to = "drug", values_to = "Cplasma") %>%
  mutate(
    drug = recode(drug,
                  Cc_rif = "Rifampicin",
                  Cc_inh = "Isoniazid",
                  Cc_emb = "Ethambutol",
                  Cc_pza = "Pyrazinamide")
  )

ggplot(sim_pk, aes(day, Cplasma, colour = drug)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~drug, scales = "free_y") +
  scale_y_continuous(name = "Plasma concentration (ug/mL)") +
  scale_x_continuous(name = "Days post first dose", breaks = c(0, 7, 14, 21, 28)) +
  labs(title = "Drug plasma trajectories under R10H25Z150E100",
       caption = "Once-daily oral gavage, 5 days on / 2 days off. Typical-value PK; no IIV.") +
  theme_minimal()
```

# Validation checks (endogenous / mechanistic)

This is a mechanistic, deterministic, typical-value model -- there is no
patient-level IIV and no PKNCA-style NCA validation. The validation strategy
follows the endogenous-model pattern (`endogenous-validation.md` in the
extraction skill).

### 1. Natural-growth check

With no drug administered, the F-state grows exponentially at rate `kG = 0.034
1/h`, while the S- and N-state populations follow from the MTP transfer
network. The total CFU/lungs should grow over 28 days from the initial
F0 + S0 = 1.39e5 to a much higher value, consistent with Chen 2017 Figure 4
"natural growth" panel.

```{r natural-growth-check}
ng_traj <- sim_obs %>% filter(regimen == "Natural growth")
ng_traj %>%
  select(day, fbugs, sbugs, nbugs, total = log10_total) %>%
  knitr::kable(digits = 2, caption = "Natural-growth trajectory of F, S, N (CFU/lungs).")
```

### 2. Antagonism check (R10H25 vs R10 + H25)

The model encodes the GPDI interaction as a slope reduction `slope_eff = slope
/ (1 + INT)`. With `INT_S_RH = 4.49`, the combined RIF + INH slope on S-state
killing is reduced to slope_mono / 5.49 -- both RIF and INH lose roughly 82% of
their S-killing potency when administered together. The day-28 CFU/lungs under
R10H25 should therefore exceed the additive-expectation CFU obtained by
combining R10 and H25 monotherapy effects.

```{r antagonism-check}
sim_obs %>%
  filter(regimen %in% c("R10 (RIF 10 mg/kg)", "H25 (INH 25 mg/kg)", "R10H25"),
         day == 28) %>%
  select(regimen, day, log10_total) %>%
  knitr::kable(digits = 3, caption = "Day-28 log10(CFU/lungs) for RIF, INH, and the R10H25 combination.")
```

### 3. Synergism check (R10H25Z150E100 includes the RIF-EMB interaction on N)

With `INT_N_RE = -0.15`, the RIF slope on N-state killing is amplified by a
factor of 1/(1 - 0.15) = 1.176 in the presence of ethambutol. Because EMB has
no quantifiable monotherapy effect, this synergistic interaction is the only
contribution EMB makes to bacterial kill in the model, leading to a further
log10(CFU/lungs) reduction at day 28 relative to R10H25Z150.

```{r synergism-check}
sim_obs %>%
  filter(regimen %in% c("R10H25Z150", "R10H25Z150E100"), day == 28) %>%
  select(regimen, day, log10_total) %>%
  knitr::kable(digits = 3, caption = "Day-28 log10(CFU/lungs) -- adding ethambutol activates the RIF-EMB synergy on N-state bacteria.")
```

# Assumptions and deviations

* **Rifampicin dose-dependent PK simplifications.** Chen 2017 reports a higher
  CL/F (94.6 mL/h/kg) and V/F (1700 mL/kg) at the highest 160 mg/kg dose, and
  a lower CL/F (54.9 mL/h/kg) in healthy mice receiving 3 weeks of treatment.
  These are not encoded structurally -- the model uses the TB-infected
  R5/R10/R20 typical values (CL/F = 234, V/F = 706) for any rifampicin dose,
  which is appropriate for the canonical 10 mg/kg combination regimen. Users
  simulating the R160 monotherapy or healthy-mouse cohorts should overwrite
  `lcl_rif` / `lvc_rif` via `mod$ini` before solving.
* **Isoniazid dose-dependent clearance.** The model implements the published
  Chen 2017 Table 1 footnote b relationship CL_inh = CL_inh_lowest *
  (1 - slope_sigma * (DOSE_INH_MGKG - 12.5)). The DOSE_INH_MGKG covariate
  carries the per-subject assigned dose level (mg/kg/day) used in the
  relationship; it is set to 0 for regimens without isoniazid.
* **GPDI functional form.** Chen 2017's Methods describes the reduced GPDI
  with EC50_INT -> 0 and joint INTAB = INTBA, but the published main-text
  formula is encoded with a placeholder ("formula-not-decoded") in the PDF
  extraction and Supplementary Material S3 (containing the NONMEM model code)
  is not on disk. The implementation here follows the canonical Wicha 2017
  GPDI formulation, slope_eff = slope_mono / (1 + INT) with INT > 0 = decreased
  potency and INT < 0 = increased potency, gated on/off by whether the
  partner drug's plasma concentration exceeds 1e-8 ug/mL. The reported
  parameter VALUES (INT_S_RH = 4.49, INT_N_RH = 0.32, INT_N_RE = -0.15) are
  encoded verbatim from Chen 2017 Table 2 and the qualitative simulation
  behaviour (antagonism in R10H25, synergism in R10H25Z150E100) matches the
  paper's published findings. Users needing perfect numerical fidelity to the
  source-paper Supplementary S3 should obtain that file and verify the
  reduction step matches this implementation.
* **Residual error scale.** Chen 2017 Table 2 reports "additive residual
  variability on log scale (variance) sigma^2 = 1.23". The MTP / GPDI
  framework (Clewe 2016, Svensson 2016) typically reports this on log10
  scale; the encoded `addSd = sqrt(1.23) * log(10)` translates to the
  natural-log scale used by nlmixr2's `add()` residual error. If the source
  paper's variance is on natural log instead (the paper text does not
  unambiguously state the log base), the appropriate value is
  `addSd = sqrt(1.23)` and users may overwrite via `mod$ini`.
* **Initial non-multiplying bacterial inoculum N0.** Chen 2017 Results section
  explicitly states "no inoculum of non-multiplying bacteria, which was,
  therefore, set to zero in the final model." Encoded literally as `nbugs(0)
  <- 0`.
* **Isoniazid effect on N-state.** kN_H was identified only in the
  R10H25Z150 triple-combination, not in isoniazid monotherapy nor R10H25
  (Chen 2017 Results). It is encoded here as a generic monotherapy slope so
  the GPDI multipliers can be applied uniformly; the simulation reproduces
  the paper's R10H25Z150 result whether INH is present alone or in a
  combination. The Discussion of Chen 2017 notes the ambiguity ("we could
  not clearly distinguish whether pyrazinamide triggered the isoniazid
  effect on N bacterial state or if this effect was an effect of isoniazid
  alone"). This deviation is the simulation-friendly choice.
* **Pyrazinamide and ethambutol monotherapy effects.** Chen 2017 Results
  notes "Due to a lack of longitudinal CFU data on ethambutol and
  pyrazinamide during monotherapy, the mono drug effects of these drugs
  could not be quantified." No monotherapy slopes are encoded for EMB or
  PZA; their contributions enter only through the RIF-EMB GPDI interaction
  on N-state (and EMB has no other effect). The PZA PK is encoded for
  completeness but does not drive any kill term.
* **5-on / 2-off dose schedule.** The simulated 4-week event table uses the
  daily-oral 5-on / 2-off schedule described in Chen 2017 Methods
  (Anti-TB treatment section). The published Figure 2 caption explicitly
  notes "The fluctuations in response are due to drug treatments in 5 of 7
  days per week", confirming this schedule is the appropriate event model
  for replication.
* **Non-canonical compartment names.** The model uses drug-suffixed
  compartments `depot_rif` / `central_rif` etc. and bacterial-state names
  `fbugs` / `sbugs` / `nbugs` (matching the Clewe 2016 in vitro MTP
  precedent). These do not match the canonical `depot` / `central` /
  `peripheral1` register and will trigger `checkModelConventions()` warnings;
  the deviation is intentional because the model carries four parallel
  drugs and three bacterial states that cannot collapse into a single
  canonical PK tree.
* **Non-canonical paper-specific covariate.** `DOSE_INH_MGKG` is a
  paper-specific covariate not in `inst/references/covariate-columns.md`.
  The canonical `DOSE` register entry assumes a single-drug context; in
  this four-drug combination model only isoniazid needs a per-subject
  dose-level covariate (its clearance is dose-dependent per Chen 2017
  Table 1 footnote b). The other three drugs have linear or fixed-dose
  PK and their assigned dose enters the simulation only through the
  event AMT column.

# Errata

* No erratum, corrigendum, or author correction has been identified for the
  Chen 2017 CPT-PSP article via the journal landing page or a PubMed
  "erratum" search as of 2026-05-17.
* Chen 2017 Supplementary Material S3 (the final NONMEM model code) is
  referenced in the paper but is not on disk in
  `from_people/literature_2018_search`. The model file extracts every
  parameter VALUE verbatim from Table 1 (popPK) and Table 2 (MTP-GPDI);
  only the precise GPDI reduction algebra is reconstructed from the
  canonical Wicha 2017 formulation per the "GPDI functional form" note
  above. A future operator with access to S3 should cross-check the
  GPDI multiplier shape.
