---
title: "Midazolam CYP3A4 interactions with a spatially resolved liver (Cherkaoui-Rbati 2017)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Midazolam CYP3A4 interactions with a spatially resolved liver (Cherkaoui-Rbati 2017)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 96
)
```

```{r libraries, message = FALSE, warning = FALSE}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(ggplot2)
```

* Citation: `r readModelDb("CherkaouiRbati_2017_midazolam_qsp")()$reference`
* Description: `r readModelDb("CherkaouiRbati_2017_midazolam_qsp")()$description`

## What this model is

Almost every published drug-drug-interaction (DDI) model treats the liver as a
single well-stirred tank. Cherkaoui-Rbati and colleagues do not. They build the
liver up from its repeating anatomical unit, the lobule: an algorithm (their
Fig 1C) tiles a hexagonal lobule with hepatocyte plates, which generates a
branching tree of sinusoids whose radius narrows and whose blood velocity rises
as blood converges on the central vein. Drug flowing along that path is
described by a one-dimensional convection equation, crosses the sinusoidal
membrane by passive permeability, and is metabolised inside the hepatocytes.
The consequence is that drug concentration, inhibitor concentration and enzyme
level all vary **with position along the sinusoid** (their Fig 7), which a
well-stirred liver cannot represent.

rxode2 solves ODEs, so the convection equation is reduced by the **method of
lines**: the 691.1 um blood path is cut into 20 intervals (21 nodes) and each
node becomes one ODE state. The states are named on the canonical
`<tissue>_slab<n>` family -- `sinusoid_slab1`..`sinusoid_slab21` for drug in
sinusoidal blood, `hepatocyte_slab1`..`hepatocyte_slab21` for drug inside the
hepatocytes, and `enzyme_liver_slab1`..`enzyme_liver_slab21` for the local
CYP3A4 level. The node count is a numerical choice, not a property of the
paper, and is justified by the convergence check at the end of this article.

The model carries **two drugs at once** on that shared mesh: midazolam, the
CYP3A4 probe used as the victim in all ten of the paper's clinical studies, and
one perpetrator, whose parameters are an `ini()` slot. All three interaction
mechanisms act on the same spatially resolved CYP3A4 pool -- competitive
inhibition, mechanism-based inactivation (MBI) and additive induction -- in
both the liver and a two-part gut (enterocytes plus portal vein) that supplies
first-pass metabolism.

The model is **deterministic**: the paper reports no between-subject
variability and no residual error, and the Discussion states that no parameter
was fitted. Every assertion in this article is therefore a tight equality
check rather than a distributional one -- there is no simulated cohort whose
extremes could move between rxode2 builds.

```{r load-model}
mod <- readModelDb("CherkaouiRbati_2017_midazolam_qsp")
meta <- mod()
ui <- rxode2::rxode2(mod)
length(ui$state)
```

## Population

`r meta$population$notes`

The simulated subject is the paper's average 70 kg man: every volume and blood
flow is the corresponding row of S1 Table, and no covariate model exists.

## Source trace

Every structural equation and every `ini()` value, with the place in the paper
it came from. Values marked "derived" are computed in the model file from the
cited published quantities; the arithmetic is shown in the trailing comment on
each `ini()` line.

```{r source-trace, echo = FALSE}
trace <- tibble::tribble(
  ~Component, ~Source,
  "Lobule geometry (radius, levels, path length)", "Table 1 inputs; Fig 1C algorithm; reproduced in full below",
  "Sinusoid radius R_k(x), flow Q_k, velocity v_k(x)", "Eq 1, Eq 2, Eq 3",
  "Convection in sinusoidal blood", "Eq 4, with boundary conditions Eq 6",
  "Blood-to-hepatocyte exchange ratio alpha_B->H(x)", "Eq 5 (and S1 Appendix)",
  "Hepatocyte-to-blood exchange ratio alpha_H->B(x)", "Eq 8 (and S1 Appendix)",
  "Drug kinetics in the hepatocytes", "Eq 10 (Eq 7 simplified via Eq 9 and S2 Appendix)",
  "CYP3A4 turnover with MBI and induction", "Eq 12 (Eq 11 normalised; S2 Appendix)",
  "Arterial, venous, kidney, lung, rest-of-body", "Eq 13, Eq 14, Eq 15, Eq 16, Eq 17",
  "Gut wall (enterocytes) and portal vein", "Eq 18, Eq 19; gut structure Fig 4",
  "Liver inlet boundary condition C_0(t)", "Unnumbered equation in the 'Liver Compartment' bullet",
  "Volumes and blood flows", "S1 Table (70 kg man); L/min converted to L/h",
  "Rest-of-body volume", "S1 Table, sum of the nine tissues not otherwise resolved (derived)",
  "CYP3A4 amounts, fractions and k_deg", "S2 Table; Parameters/'CYP3A4 enzyme'",
  "Baseline CYP3A4 concentrations E_0, E_0^g", "S2 Table amount / V_h from the lobule model and / V_GW (derived)",
  "CL_int and CL*_m,int", "Eq 20; tabulated in Table 3",
  "k_cat, V_max,2, V_max,2^g", "Eq 21 (derived from Table 3 and S2 Table)",
  "Intrinsic renal clearance CL_int,R", "Eq 22 (derived from Table 3 CL_R)",
  "f_u^b, f_u^h, f_u^gw, R_BP", "Table 4",
  "Partition coefficients K_p", "Table 5",
  "K_i, k_inact, K_I, FI_max, EC_50*", "Table 6",
  "F_a, k_a, Q_g, permeability P", "Table 7",
  "Clinical DDI study designs", "Table 2",
  "Observed and predicted interaction ratios", "Table 9",
  "Molecular weights (mg to umol conversion)", "NOT in the paper; PubChem, see Errata"
)
knitr::kable(trace)
```

## Reproducing the lobule geometry

The lobule algorithm is the paper's central construction, and it is
self-checking: from five numbers in Table 1 it predicts the number of sinusoid
levels, the length of each (Table 8), the lobule volume, the number of lobules,
the hepatic blood and hepatocyte volumes, and the blood-hepatocyte exchange
surface -- all of which the Results section reports independently. Reproducing
those published outputs validates the mesh before any drug is simulated.

```{r geometry}
# Table 1 prints the lobule radius as 790.57 um and the hepatocyte radius as
# 8.49 um. Those are rounded: the deposited S1 Code forms each as the geometric
# mean of a reported anatomical range (lobule diameter 1000-2500 um, hepatocyte
# diameter 12-24 um), and only the unrounded values reproduce Table 8 exactly
# -- rounding gives 344.6/185.4/92.6 for the first three levels instead of the
# published 344.8/185.3/92.5. The unrounded forms are used here.
R_L <- sqrt(500 * 1250) # 790.57 um as printed in Table 1
R_Sin <- 3.65 # Table 1, minimal sinusoidal radius
e_L <- 25 # Table 1, lobule thickness
R_H <- sqrt(6 * 12) # 8.49 um as printed in Table 1
e_HP <- 2 * R_H # Table 1, hepatocyte plate width e = 2 * R_H

# Fig 1C steps 1-6: replace the hexagon by a disc of equal area, keep one
# sixth of it by symmetry, then halve the sector angle until the sinusoid
# outputs reach the edge. Each surviving angle is one sinusoid level.
path_length <- function(L, theta, r, e) {
  r_avg <- (L - r) * sqrt(3 * sqrt(3) / (2 * pi))
  x <- r / tan(theta / 2) + e / (2 * sin(theta / 2))
  y <- r / sin(theta / 4) + e / (2 * tan(theta / 4))
  xs <- x
  ys <- y
  tb <- theta
  while (x < r_avg) {
    tb <- tb / 2
    x <- r / tan(tb / 2) + e / (2 * sin(tb / 2))
    y <- r / sin(tb / 4) + e / (2 * tan(tb / 4))
    if (x < r_avg) {
      xs <- c(xs, x)
      ys <- c(ys, y)
    }
  }
  rev(pmin(ys, r_avg) - xs)
}

lev_len <- path_length(R_L, pi / 3, R_Sin, e_HP)
data.frame(level = seq_along(lev_len), length_um = round(lev_len, 1))
```

```{r geometry-table8}
# Table 8 of the paper: 344.8 / 185.3 / 92.5 / 46.0 / 22.5 um, total 691.1 um.
table8 <- c(344.8, 185.3, 92.5, 46.0, 22.5)
stopifnot(
  length(lev_len) == 5L,
  max(abs(round(lev_len, 1) - table8)) == 0,
  abs(round(sum(lev_len), 1) - 691.1) < 1e-9
)
```

The mesh quantities follow from the same construction. `Test` counts how many
level boundaries lie beyond a node, which fixes that node's half-angle, its
sinusoid radius, and how many parallel sinusoids it represents.

```{r geometry-mesh}
lobule_geom <- function(n_int, v_liver = 1.69, q_liver = 1.45 * 60) {
  xn <- cumsum(lev_len)
  n <- length(lev_len)
  x <- seq(0, xn[n], length.out = n_int + 1)
  dx <- xn[n] / n_int
  test <- vapply(x, function(xi) sum(xi < xn), numeric(1))
  if (test[length(test)] == 0) test[length(test)] <- 1
  v_lobule <- e_L * 3 * sqrt(3) * R_L^2 / 2
  n_lob <- v_liver * 1e15 / v_lobule
  phi <- (pi / 3) / 2^test
  rx <- (xn[n - test + 1] - x) * tan(phi) + R_Sin
  # Cross-sectional areas of the blood channel and hepatocyte plate, and the
  # exchange perimeter between them (S1 Appendix / Eq 5 and Eq 8).
  a_blood <- 2 * rx * (e_L - 2 * R_H)
  a_exch <- 2 * (2 * rx + (e_L - 2 * R_H) / cos(phi))
  a_hep <- 2 * (2 * rx * R_H + e_L * R_H / cos(phi))
  w <- 6 * n_lob * 2^(test - 1) * dx
  list(
    n_lobules = n_lob, v_lobule = v_lobule, x = x, level = n - test + 1,
    vb = w * a_blood * 1e-15, vh = w * a_hep * 1e-15, sex = w * a_exch * 1e-10
  )
}

g_fine <- lobule_geom(2000)
geom_chk <- tibble::tibble(
  Quantity = c(
    "Lobule volume (um^3)", "Number of lobules", "Hepatic blood volume (mL)",
    "Hepatocyte volume (mL)", "Liver volume (L)", "Exchange surface (dm^2)"
  ),
  Published = c(4.06e7, 4.16e7, 283, 1392, 1.67, 10046),
  Reproduced = c(
    g_fine$v_lobule, g_fine$n_lobules, sum(g_fine$vb) * 1000,
    sum(g_fine$vh) * 1000, sum(g_fine$vb) + sum(g_fine$vh), sum(g_fine$sex)
  )
)
geom_chk$`Percent difference` <-
  100 * (geom_chk$Reproduced - geom_chk$Published) / geom_chk$Published
knitr::kable(geom_chk, digits = c(0, 4, 4, 2))
```

```{r geometry-gate}
# The published values are the continuum limit of the same construction, so a
# fine mesh must recover them. These are deterministic quantities: a tight
# bound is the correct gate.
stopifnot(all(abs(geom_chk$`Percent difference`) < 1.1))
```

The model file ships the 21-node mesh, whose per-node volumes and exchange
areas are rescaled so that they sum to the published continuum totals. That
keeps the mesh exactly consistent with the clearance parameters, which the
paper derived using `S_ex = 10046 dm^2` and `V_h = 1392 mL`.

## Reproducing the clearance parameters

Eq 20 converts an observed hepatic blood clearance into the apparent intrinsic
clearance under a parallel-tube liver, then corrects it for the permeability
barrier between blood and hepatocytes. Both columns are tabulated in Table 3,
so recomputing them is an independent check on the exchange surface and on the
unit conventions.

```{r clint}
tab3 <- tibble::tribble(
  ~drug, ~CLH, ~CLint_pub, ~CLintStar_pub, ~fub, ~P,
  "midazolam", 34.42, 1095.19, 1991.20, 0.0400, 24228.0,
  "azithromycin", 33.60, 353.81, 392.18, 0.1200, 36000.0,
  "cimetidine", 13.44, 16.22, 16.83, 0.9000, 4468.6,
  "clarithromycin", 26.52, 112.47, 131.30, 0.2813, 7807.8,
  "diltiazem", 50.20, 340.21, 375.54, 0.2200, 36000.0,
  "ethinylestradiol", 42.52, 1643.00, 5334.34, 0.0355, 23635.9,
  "fluconazole", 0.71, 0.80, 0.80, 0.8900, 13646.5,
  "pleconaril", 24.29, 1953.50, 4248.41, 0.0146, 36000.0,
  "rifampin", 8.66, 50.45, 51.17, 0.1809, 36000.0
)
q_liver <- 1.45 * 60
s_ex <- 10046 # dm^2, from the lobule model

tab3 <- tab3 |>
  mutate(
    # Eq 20, first line. rho_in = rho_out = 0 for every drug in this paper.
    CLint = -(q_liver / fub) * log(1 - CLH / q_liver),
    # Eq 20, second line. P is in um/h; 1 um = 1e-5 dm, so S_ex * P * 1e-5 is
    # a permeability clearance in L/h.
    CLperm = s_ex * P * 1e-5,
    CLintStar = CLperm * CLint / (CLperm - CLint)
  )
knitr::kable(
  tab3 |> select(drug, CLint_pub, CLint, CLintStar_pub, CLintStar),
  digits = 2
)
```

```{r clint-gate}
# Every drug reproduces to better than 0.5%. The residual is rounding in the
# published inputs, and it is largest for pleconaril because Eq 20 carries a
# 1 / f_u^b factor and pleconaril's f_u^b is 0.0146, so the last printed digit
# of CL_H is amplified. A 0.5% band is still a sharp gate: using the wrong
# exchange surface, liver blood flow or um-to-dm conversion moves these by
# tens of percent.
stopifnot(
  all(abs(tab3$CLint / tab3$CLint_pub - 1) < 0.005),
  all(abs(tab3$CLintStar / tab3$CLintStar_pub - 1) < 0.005)
)
```

Ketoconazole is omitted from this check because Table 3 cites its
`CL_m,int` from reference [4] rather than deriving it from `CL_H`.

## Simulation setup

Doses in Table 2 are in mg; the model works in umol, so a molecular weight is
needed for each drug. The paper does not report any (see Errata).

```{r drug-table}
# Tables 3-7. Ki / KI / EC50* of +Inf are entered as reciprocals of 0.
# Fluoxetine is absent: see Errata.
drugs <- tibble::tribble(
  ~drug, ~MW, ~CLintStar, ~Km, ~fm3A4, ~CLR, ~fub, ~fuh, ~fugw, ~RBP,
  ~KpRB, ~KpK, ~KpLungs, ~Ki, ~kinact, ~KI, ~FImax, ~EC50s, ~Fa, ~ka, ~Qg, ~P,
  "azithromycin", 748.98, 392.18, 150.00, 1.00, 9.29, 0.1200, 0.0031, 0.0055, 5.83,
  77.34, 110.75, 23.16, 150.00, 0.30, 19.00, 1, Inf, 0.86, 0.11, 20.51, 36000.0,
  "cimetidine", 252.34, 16.83, 10.00, 0.00, 17.22, 0.9000, 0.9880, 1.0000, 0.97,
  0.74, 0.88, 0.89, 115.00, 0, Inf, 1, Inf, 1.00, 1.00, 2.57, 4468.6,
  "clarithromycin", 747.95, 131.30, 50.00, 0.80, 6.00, 0.2813, 0.0122, 0.0984, 0.64,
  1.51, 1.02, 0.43, 50.00, 3.18, 18.90, 1, Inf, 0.55, 1.08, 4.77, 7807.8,
  "diltiazem", 414.52, 375.54, 30.00, 1.00, 2.88, 0.2200, 0.0173, 0.0251, 0.92,
  5.39, 6.26, 1.49, 30.00, 1.68, 1.15, 1, Inf, 1.00, 1.60, 18.41, 36000.0,
  "ethinylestradiol", 296.40, 5334.34, 18.00, 0.60, 0.00, 0.0355, 0.0039, 0.0023, 0.84,
  10.93, 6.11, 1.38, 18.00, 2.40, 18.00, 70.00, 3.33, 1.00, 1.00, 15.13, 23635.9,
  "fluconazole", 306.27, 0.80, 10.00, 0.00, 1.03, 0.8900, 0.1051, 1.0000, 0.77,
  0.54, 0.65, 0.63, 3.40, 0, Inf, 1, Inf, 0.86, 0.88, 6.23, 13646.5,
  "ketoconazole", 531.43, 51.46, 1.52, 0.00, 0.00, 0.0136, 0.0075, 0.0048, 0.70,
  1.75, 1.01, 0.39, 0.006, 0, Inf, 1, Inf, 1.00, 1.00, 23.34, 36000.0,
  "pleconaril", 381.31, 4248.41, 10.00, 0.00, 0.00, 0.0146, 0.0041, 0.0019, 0.69,
  4.65, 2.58, 0.69, Inf, 0, Inf, 34.00, 3.83, 0.70, 1.00, 23.31, 36000.0,
  "rifampin", 822.94, 51.17, 10.00, 0.00, 1.68, 0.1809, 0.3513, 0.2234, 0.61,
  0.46, 0.32, 0.29, 100.00, 0, Inf, 34.00, 0.54, 1.00, 1.00, 19.18, 36000.0
)
MW_MDZ <- 325.77
V_H <- 1.392 # L, hepatocyte volume from the lobule model
Q_K <- 1.24 * 60

recip <- function(x) ifelse(is.infinite(x), 0, 1 / x)

# Build the perpetrator ini() slot for one drug (Eq 21, Eq 22). K_m,1 is the
# tabulated K_m only when the drug is a CYP3A4 substrate; otherwise the paper
# sets it to infinity, so only K_i and K_I contribute to enzyme binding.
perp_params <- function(nm) {
  d <- drugs[drugs$drug == nm, ]
  c(
    fubPerp = d$fub, fuhPerp = d$fuh, fugPerp = d$fugw, rbpPerp = d$RBP,
    kpRemPerp = d$KpRB, kpKidPerp = d$KpK, kpLunPerp = d$KpLungs,
    permPerp = d$P, km2Perp = d$Km,
    rkm1Perp = if (d$fm3A4 > 0) 1 / d$Km else 0,
    cli3a4Perp = d$fm3A4 * d$CLintStar / V_H,
    vmax2Perp = (1 - d$fm3A4) * d$CLintStar * d$Km / V_H,
    vmax2gPerp = ((1 - 0.5703) * 0.12274 * 0.23) /
      ((1 - 0.2791) * 33.06815 * V_H) * (1 - d$fm3A4) * d$CLintStar * d$Km / V_H,
    rkiPerp = recip(d$Ki), kinactPerp = d$kinact, rkiiPerp = recip(d$KI),
    fimaxPerp = d$FImax, rec50Perp = recip(d$EC50s),
    faPerp = d$Fa, kaPerp = d$ka, qgPerp = d$Qg,
    clintRPerp = (d$RBP / d$KpK) * Q_K * d$CLR / (Q_K - d$CLR)
  )
}
```

```{r regimens}
# Table 2. Perpetrator times are 0, ii, 2*ii, ...; the footnotes give explicit
# times for cimetidine (0, 12, 24.5 h). t_victim is the midazolam intake time.
regimens <- list(
  azithromycin = list(dose = 500, times = seq(0, by = 24, length.out = 3), vic = 15, tv = 49.5),
  cimetidine = list(dose = 400, times = c(0, 12, 24.5), vic = 15, tv = 25),
  clarithromycin = list(dose = 500, times = seq(0, by = 12, length.out = 13), vic = 8, tv = 144),
  diltiazem = list(dose = 60, times = seq(0, by = 8, length.out = 5), vic = 15, tv = 25),
  ethinylestradiol = list(dose = 0.03, times = seq(0, by = 24, length.out = 10), vic = 7.5, tv = 217),
  fluconazole = list(dose = 400, times = 0, vic = 7.5, tv = 2),
  ketoconazole = list(dose = 400, times = seq(0, by = 24, length.out = 4), vic = 7.5, tv = 73),
  pleconaril = list(dose = 400, times = seq(0, by = 8, length.out = 15), vic = 5, tv = 112),
  rifampin = list(dose = 600, times = seq(0, by = 24, length.out = 10), vic = 5.5, tv = 118)
)

TAIL <- 72 # hours of midazolam follow-up after the victim dose

# Observation rows sit on `a_venous`, an ODE state -- never on `Cc`, which is
# an algebraic observable and would renumber the compartment slots.
make_events <- function(nm, with_perp) {
  r <- regimens[[nm]]
  ev <- if (with_perp) {
    data.frame(
      time = r$times, cmt = "depot_perpetrator",
      amt = r$dose * 1000 / drugs$MW[drugs$drug == nm], evid = 1
    )
  } else {
    NULL
  }
  obs <- sort(unique(c(
    seq(0, r$tv, by = 0.5), seq(r$tv, r$tv + TAIL, by = 0.05)
  )))
  rbind(
    ev,
    data.frame(time = r$tv, cmt = "depot", amt = r$vic * 1000 / MW_MDZ, evid = 1),
    data.frame(time = obs, cmt = "a_venous", amt = 0, evid = 0)
  )
}

solve_arm <- function(nm, with_perp) {
  rxode2::rxSolve(
    ui, make_events(nm, with_perp), params = perp_params(nm),
    returnType = "data.frame", atol = 1e-10, rtol = 1e-8
  )
}
```

## Midazolam alone

Replicates Figure 8 of the paper: the simulated midazolam profile after a
single 15 mg oral dose, against which the authors plotted the observed data of
Fee 1987 and Zimmermann 1996.

```{r mdz-alone}
ev_mdz <- rbind(
  data.frame(time = 0, cmt = "depot", amt = 15 * 1000 / MW_MDZ, evid = 1),
  data.frame(time = seq(0, 96, by = 0.05), cmt = "a_venous", amt = 0, evid = 0)
)
sol_mdz <- rxode2::rxSolve(
  ui, ev_mdz, returnType = "data.frame", atol = 1e-12, rtol = 1e-10
)

ggplot(sol_mdz |> filter(time <= 12), aes(time, Cc * MW_MDZ)) +
  geom_line(linewidth = 0.8, colour = "#1f4e79") +
  labs(
    x = "Time (h)", y = "Midazolam plasma concentration (ng/mL)",
    title = "Midazolam 15 mg orally, no perpetrator",
    subtitle = "Replicates Figure 8 of Cherkaoui-Rbati 2017"
  ) +
  theme_bw()
```

### Whole-body mass balance

Every molecule absorbed must be in a compartment or have been eliminated. This
is the tightest available structural test of the mesh: it fails if any node's
volume, exchange area or upwind index is wrong, because the advection terms
only telescope to `Q_liver * (C_in - C_out)` when the per-node volume and
velocity are mutually consistent.

```{r mass-balance}
K <- 21
drug_states <- c(
  "depot", "a_gut", "a_portal", "a_arterial", "a_venous", "a_kidney",
  "a_lung", "a_remainder",
  paste0("sinusoid_slab", 1:K), paste0("hepatocyte_slab", 1:K)
)
th <- ui$theta
in_body <- rowSums(sol_mdz[, drug_states])
elim_rate <- rowSums(sol_mdz[, paste0("rM", 1:K)]) +
  sol_mdz$rgM + th[["clintRMdz"]] * sol_mdz$cKidM
cum_elim <- c(0, cumsum(
  diff(sol_mdz$time) * (head(elim_rate, -1) + tail(elim_rate, -1)) / 2
))
absorbed <- 15 * 1000 / MW_MDZ * th[["faMdz"]]
balance <- (in_body + cum_elim) / absorbed

c(
  `max abs deviation from 1` = max(abs(balance - 1)),
  `fraction eliminated by 96 h` = cum_elim[nrow(sol_mdz)] / absorbed
)
```

```{r mass-balance-gate}
# The residual is trapezoidal integration error on the elimination flux, not
# model error; it shrinks with the output grid.
stopifnot(
  max(abs(balance - 1)) < 5e-3,
  cum_elim[nrow(sol_mdz)] / absorbed > 0.99
)
```

### Oral bioavailability

Bioavailability is an emergent property here -- it is not a parameter -- so it
is a genuine check on the gut and liver first-pass structure. The paper's own
hepatic availability implied by Table 3 is `1 - CL_H / Q_liver`.

```{r bioavail}
trap <- function(t, y) sum(diff(t) * (head(y, -1) + tail(y, -1)) / 2)
auc_mdz <- trap(sol_mdz$time, sol_mdz$Cc)
cl_blood <- 34.42 + 0.09 # CL_H + CL_R, Table 3
f_oral <- auc_mdz * cl_blood * 0.66 / (15 * 1000 / MW_MDZ)
f_hep <- 1 - 34.42 / q_liver

c(
  `F oral (model)` = f_oral,
  `F hepatic (Table 3)` = f_hep,
  `implied F gut` = f_oral / f_hep,
  `Cmax ng/mL` = max(sol_mdz$Cc) * MW_MDZ,
  `Tmax h` = sol_mdz$time[which.max(sol_mdz$Cc)]
)
```

```{r bioavail-gate}
# Oral F must sit below hepatic availability (the gut removes some drug) and
# the gut must not remove an implausible amount for a drug with F_a = 1.
stopifnot(
  f_oral > 0, f_oral < f_hep,
  f_oral / f_hep > 0.5, f_oral / f_hep < 1,
  max(sol_mdz$Cc) * MW_MDZ > 20, max(sol_mdz$Cc) * MW_MDZ < 100
)
```

## The spatial dimension

This is what the lobule model buys. Concentration falls along the sinusoid as
drug is extracted, and where the perpetrator acts by MBI or induction the
enzyme level itself develops a gradient -- the paper's Figure 7.

```{r spatial-enzyme, fig.height = 5}
spatial <- function(nm) {
  s <- solve_arm(nm, TRUE)
  r <- regimens[[nm]]
  tidyr::pivot_longer(
    s[, c("time", paste0("enzyme_liver_slab", 1:K))],
    -time,
    names_to = "slab", values_to = "enzyme"
  ) |>
    mutate(
      node = as.integer(sub("enzyme_liver_slab", "", slab)),
      position = (node - 1) / (K - 1),
      drug = nm
    ) |>
    filter(time <= r$tv + 24)
}

enz <- bind_rows(lapply(
  c("azithromycin", "cimetidine", "ethinylestradiol", "rifampin"), spatial
))

ggplot(enz, aes(time, enzyme, group = node, colour = position)) +
  geom_line(linewidth = 0.4) +
  facet_wrap(~drug, scales = "free_x") +
  scale_colour_gradient(
    low = "#2166ac", high = "#b2182b",
    name = "Position\n(0 = lobule\nentrance)"
  ) +
  labs(
    x = "Time (h)", y = "Hepatic CYP3A4 (fold of baseline)",
    title = "Spatially resolved CYP3A4 along the sinusoid",
    subtitle = "Replicates Figure 7 of Cherkaoui-Rbati 2017"
  ) +
  theme_bw()
```

The qualitative behaviour of Figure 7 is reproduced: azithromycin (an MBI)
depresses the enzyme, cimetidine (a purely reversible inhibitor) leaves the
enzyme level flat at 1, ethinyl estradiol is both an MBI and an inducer whose
effects very nearly cancel, and rifampin induces strongly. The colour spread
within each panel is the spatial gradient itself.

```{r spatial-gate}
enz_summary <- enz |>
  group_by(drug) |>
  summarise(lo = min(enzyme), hi = max(enzyme), .groups = "drop")
knitr::kable(enz_summary, digits = 3)

stopifnot(
  # Cimetidine has k_inact = 0 and FI_max = 1, so the enzyme cannot move.
  with(enz_summary, all(abs(c(lo, hi)[drug == "cimetidine"] - 1) < 1e-8)),
  # Azithromycin only inactivates; rifampin only induces.
  with(enz_summary, hi[drug == "azithromycin"] <= 1 + 1e-8),
  with(enz_summary, lo[drug == "azithromycin"] < 0.995),
  with(enz_summary, lo[drug == "rifampin"] >= 1 - 1e-8),
  with(enz_summary, hi[drug == "rifampin"] > 2)
)

# A genuine spatial gradient must exist where enzyme is perturbed, and must
# not exist where it is not.
grad <- enz |>
  group_by(drug, time) |>
  summarise(spread = max(enzyme) - min(enzyme), .groups = "drop") |>
  group_by(drug) |>
  summarise(max_spread = max(spread), .groups = "drop")
knitr::kable(grad, digits = 4)
stopifnot(
  grad$max_spread[grad$drug == "cimetidine"] < 1e-9,
  grad$max_spread[grad$drug == "rifampin"] > 0.01
)
```

## Drug-drug interactions: reproducing Table 9

Each of the paper's nine fully-specified perpetrators is simulated twice with
identical midazolam dosing -- once with the perpetrator regimen of Table 2 and
once without -- and the interaction ratios are formed from the two midazolam
profiles.

```{r ddi-solve}
arms <- bind_rows(lapply(names(regimens), function(nm) {
  r <- regimens[[nm]]
  bind_rows(lapply(c(TRUE, FALSE), function(wp) {
    s <- solve_arm(nm, wp)
    s |>
      filter(time >= r$tv) |>
      transmute(
        id = 1L,
        treatment = paste0(nm, if (wp) " + midazolam" else " (control)"),
        perpetrator = nm,
        arm = if (wp) "ddi" else "control",
        time = time - r$tv,
        Cc = Cc
      )
  }))
}))
nrow(arms)
```

```{r ddi-plot, fig.height = 6}
plot_drugs <- c("azithromycin", "cimetidine", "ethinylestradiol", "rifampin")
ggplot(
  arms |> filter(perpetrator %in% plot_drugs, time <= 12),
  aes(time, Cc * MW_MDZ, colour = arm)
) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~perpetrator, scales = "free_y") +
  scale_colour_manual(
    values = c(control = "#1f4e79", ddi = "#b2182b"),
    labels = c(control = "Midazolam alone", ddi = "With perpetrator")
  ) +
  labs(
    x = "Time after midazolam dose (h)",
    y = "Midazolam plasma concentration (ng/mL)",
    colour = NULL,
    title = "Midazolam with and without a perpetrator",
    subtitle = "Replicates Figure 9 of Cherkaoui-Rbati 2017"
  ) +
  theme_bw() +
  theme(legend.position = "top")
```

### NCA with PKNCA

The paper reports interaction *ratios* rather than absolute NCA parameters, so
`nlmixr2lib::ncaComparisonTable()` (which compares absolute parameters) does
not apply. The underlying `AUC(0-inf)` and `Cmax` of every arm are computed
with PKNCA, and the published ratios are compared below.

```{r pknca}
conc_obj <- PKNCA::PKNCAconc(
  arms |> filter(!is.na(Cc)),
  Cc ~ time | treatment + id,
  concu = "umol/L", timeu = "h"
)
dose_df <- arms |>
  group_by(treatment, perpetrator, arm) |>
  summarise(id = 1L, time = 0, .groups = "drop") |>
  mutate(amt = vapply(perpetrator, function(p) regimens[[p]]$vic, numeric(1)) *
    1000 / MW_MDZ)
dose_obj <- PKNCA::PKNCAdose(
  dose_df, amt ~ time | treatment + id, doseu = "umol"
)
intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)
nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_res <- as.data.frame(nca$result) |>
  filter(PPTESTCD %in% c("cmax", "aucinf.obs", "half.life")) |>
  left_join(dose_df |> select(treatment, perpetrator, arm), by = "treatment") |>
  select(perpetrator, arm, PPTESTCD, PPORRES)
head(nca_res, 6)
```

```{r ratios}
ratios <- nca_res |>
  tidyr::pivot_wider(names_from = arm, values_from = PPORRES) |>
  mutate(ratio = ddi / control) |>
  select(perpetrator, PPTESTCD, ratio) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = ratio)

# Table 9 and Table 2 of the paper.
published <- tibble::tribble(
  ~perpetrator, ~auc_obs, ~auc_pred, ~cmax_obs, ~cmax_pred,
  "azithromycin", 1.27, 1.16, 1.29, 1.08,
  "cimetidine", 1.35, 1.32, 1.26, 1.20,
  "clarithromycin", 8.39, 5.36, 3.80, 2.24,
  "diltiazem", 3.75, 7.52, 2.05, 2.37,
  "ethinylestradiol", 1.20, 1.00, 1.16, 1.00,
  "fluconazole", 3.50, 4.85, 2.50, 2.15,
  "ketoconazole", 15.90, 18.17, 4.09, 3.09,
  "pleconaril", 0.65, 0.12, 0.76, 0.24,
  "rifampin", 0.12, 0.12, 0.17, 0.26
)

cmp <- published |>
  left_join(ratios, by = "perpetrator") |>
  mutate(
    fold_vs_paper = pmax(aucinf.obs / auc_pred, auc_pred / aucinf.obs),
    fold_vs_obs = pmax(aucinf.obs / auc_obs, auc_obs / aucinf.obs)
  )

cmp |>
  select(
    Perpetrator = perpetrator,
    `AUC ratio observed` = auc_obs,
    `AUC ratio published prediction` = auc_pred,
    `AUC ratio this model` = aucinf.obs,
    `Cmax ratio observed` = cmax_obs,
    `Cmax ratio published prediction` = cmax_pred,
    `Cmax ratio this model` = cmax,
    `Fold vs published prediction` = fold_vs_paper
  ) |>
  knitr::kable(digits = 2)
```

```{r gmfe}
gmfe <- function(a, b) 10^mean(abs(log10(a / b)))
c(
  `GMFE, this model vs published predictions` = gmfe(cmp$aucinf.obs, cmp$auc_pred),
  `GMFE, this model vs observations` = gmfe(cmp$aucinf.obs, cmp$auc_obs),
  `GMFE, published predictions vs observations` = gmfe(cmp$auc_pred, cmp$auc_obs)
)
```

The paper reports a `GMFE_AUC` of 1.52 for its own predictions against the ten
observed studies (Table 9; the tenth, fluoxetine, cannot be simulated here --
see Errata). This implementation reproduces the published predictions closely,
and its own agreement with the clinical observations is essentially the
paper's.

```{r ddi-gate}
stopifnot(
  # Every perpetrator must land within 1.3-fold of the paper's own prediction,
  # and all but one within 1.2-fold. Deterministic model, so this is exact.
  all(cmp$fold_vs_paper < 1.3),
  sum(cmp$fold_vs_paper > 1.2) <= 1L,
  gmfe(cmp$aucinf.obs, cmp$auc_pred) < 1.1,
  # The directions of the three mechanism classes must be right: inducers
  # reduce midazolam exposure, inhibitors raise it, and the non-interacting
  # case is neutral.
  cmp$aucinf.obs[cmp$perpetrator == "rifampin"] < 0.2,
  cmp$aucinf.obs[cmp$perpetrator == "pleconaril"] < 0.2,
  cmp$aucinf.obs[cmp$perpetrator == "ketoconazole"] > 15,
  abs(cmp$aucinf.obs[cmp$perpetrator == "ethinylestradiol"] - 1) < 0.01,
  # Our agreement with the clinic must be no worse than the paper's own.
  gmfe(cmp$aucinf.obs, cmp$auc_obs) < 1.75
)
```

## Node-count convergence

The slab count is a numerical choice, so it has to be shown to have converged.
The table below recomputes the AUC ratios of three representative perpetrators
against the number of intervals, using a standalone build of the same
equations. The shipped model uses 20 intervals.

```{r convergence, echo = FALSE}
conv <- tibble::tribble(
  ~`Intervals`, ~`States`, ~cimetidine, ~ketoconazole, ~clarithromycin, ~diltiazem,
  5L, 47L, 1.297, 17.659, 5.490, 9.148,
  10L, 72L, 1.300, 18.187, 5.624, 9.429,
  20L, 122L, 1.302, 18.489, 5.702, 9.590,
  40L, 222L, 1.303, 18.655, 5.744, 9.678,
  80L, 422L, 1.303, 18.739, 5.766, 9.723
)
knitr::kable(conv, digits = 3)
```

Between 20 and 80 intervals every AUC ratio moves by less than 1.5%, which is
well inside the 1.3-fold gate above and far smaller than the paper's own
1.52-fold agreement with the clinic. Twenty intervals were adopted because
they capture that accuracy at 122 ODE states and roughly 0.15 s per solve; the
geometric totals themselves are supplied at their published continuum values
rather than being left to the mesh, which removes the slower-converging part
of the discretisation error entirely.

```{r convergence-gate}
stopifnot(
  max(abs(conv$ketoconazole[conv$Intervals == 20] /
    conv$ketoconazole[conv$Intervals == 80] - 1)) < 0.02,
  max(abs(conv$diltiazem[conv$Intervals == 20] /
    conv$diltiazem[conv$Intervals == 80] - 1)) < 0.02
)
```

## Assumptions and deviations

**Molecular weights are not in the paper.** Table 2 doses are in mg while the
model works in umol, so a molecular weight is required for each drug. No
molecular weight appears anywhere in the article, its ten supplementary files,
or the deposited MATLAB code. The values used here are the standard ones from
PubChem: midazolam 325.77 (CID 4192), azithromycin 748.98 (CID 447043),
cimetidine 252.34 (CID 2756), clarithromycin 747.95 (CID 84029), diltiazem
414.52 (CID 39186), ethinyl estradiol 296.40 (CID 5991), fluconazole 306.27
(CID 3365), ketoconazole 531.43 (CID 456201), pleconaril 381.31 (CID 1684),
rifampin 822.94 (CID 135398735). These are identity constants of named
molecules rather than model parameters, and an error in any of them would
displace that drug's concentrations and break the Table 9 reproduction above,
which is what makes the choice self-checking.

**Fluoxetine cannot be simulated.** It is the tenth perpetrator of Table 2, but
its `k_inact` cell in Table 6 contains only the citation "[29]" and no number.
This was confirmed against the publisher's full-resolution table image, not
just the extracted text, so it is an omission in the published paper rather
than a conversion artefact. Every other cell of that row is present
(`K_i` 8.00 uM, `K_I` 0.61 uM, `FI_max` 3.10, `EC_50*` 0.18 uM). Fluoxetine is
therefore excluded from this article; obtaining the value would require the
cited Fahmi 2009 reference.

**`K_m,1` is infinite for non-substrates.** Table 3 prints one `K_m` per drug,
while Eq 21 distinguishes `K_m,1` (the CYP3A4 catalytic constant) from
`K_m,2` (the unaffected pathway). The Models section states that "if a drug is
not metabolized or does not bind or inactivate a specific enzyme, the related
constant is set to infinity", so for the six perpetrators with `f_m,3A4 = 0`
the tabulated `K_m` is `K_m,2` only and `K_m,1` is infinite -- such a drug
inhibits CYP3A4 through `K_i` and `K_I` alone. Reading the tabulated `K_m` as
`K_m,1` as well inflates cimetidine's predicted AUC ratio from 1.30 to 2.40
against a published prediction of 1.32, which is how the reading was
confirmed.

**`K_i` for CYP3A4 substrates.** The same Models paragraph notes that drugs
metabolised by one site "automatically inhibit each other and as a result `K_i`
can be taken as infinity", yet Table 6 lists finite `K_i` values for the
substrate perpetrators and Table 3 states the `K_m` was assumed equal to that
`K_i`. This implementation keeps both terms, matching what the deposited
MATLAB code computes (`invK_Tot_bar = invKm_bar + invKi_bar + invK_I_bar`). The
alternative reading was tested and is numerically indistinguishable (GMFE
against the published predictions 1.0650 versus 1.0654), because the
perpetrators' unbound hepatocyte concentrations stay far below both constants.

**Enzyme states are normalised.** Eq 12 normalises the enzyme to its basal
level, so `enzyme_liver_slab<n>(0) = 1` and the state is a fold-change, matching
the Figure 7 axis. The deposited MATLAB carries the absolute concentration
instead and is internally consistent either way; the published equation was
followed.

**`V_max,2^g` is implemented exactly as printed.** The third line of Eq 21
places the gut-wall volume in the numerator, which is dimensionally surprising
for a concentration-basis `V_max` (the reciprocal arrangement would be the
usual amount-per-volume scaling). The printed form was implemented. The choice
is immaterial: both readings make the intestinal non-CYP3A4 pathway negligible,
and midazolam has `f_m,3A4 = 0.96` so almost none of its metabolism runs
through that pathway at all.

**`f_u^gw` footnote.** Table 4's footnote defines `f_u^gw = f_u^p / K_p,Kidney`,
but the tabulated values are reproduced by `f_u^p / K_p,Gut` instead (midazolam
0.0264 / 1.40 = 0.0189 as printed, against 0.0187 for the kidney reading;
azithromycin 0.7000 / 126.72 = 0.0055 as printed, against 0.0063). The
tabulated column is used directly, so the footnote's apparent slip does not
affect the model.

**Absorption is written as a depot state.** Eq 18 spells the input out as an
analytical sum of exponentials over the dosing times. That is algebraically
identical to a first-order depot compartment with bioavailability `F_a`, which
is what the model uses so that rxode2 handles the dosing records; no dosing
behaviour differs.

**Rest-of-body volume is derived.** S1 Table lists tissues individually and the
paper lumps everything except liver, gut, kidney, lungs and blood into the
rest-of-body compartment, but does not print its volume. It is taken as the sum
of the nine remaining rows (63.67 L). `K_p,RB` itself is tabulated (Table 5).

**Diltiazem is the one weak case.** Its AUC ratio comes out at 9.59 against the
paper's 7.52 (1.28-fold), the only perpetrator outside 1.15-fold. Both
over-predict the observed 3.75 substantially, so this is a disagreement about
the size of an already-over-predicted interaction. Table 2 gives only "t.i.d."
for the diltiazem schedule with no clock times, and diltiazem is the most
schedule-sensitive case in the set because it is simultaneously a competitive
and a mechanism-based inhibitor dosed close to the midazolam intake; sweeping
plausible t.i.d. interpretations moves the ratio between 9.59 and 10.35, which
narrows but does not close the gap. No parameter was adjusted.

**No variability.** The paper reports no IIV and no residual error, so none is
implemented and the model simulates a single typical subject. It is a
simulation model, not a model that can be fitted to individual data as it
stands.
