---
title: "DSM265 malaria chemoprophylaxis PK/PD (Cherkaoui-Rbati 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DSM265 malaria chemoprophylaxis PK/PD (Cherkaoui-Rbati 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("CherkaouiRbati_2023_dsm265"))$reference`
* Description: `r rxode2::rxode(readModelDb("CherkaouiRbati_2023_dsm265"))$description`
* Article: <https://doi.org/10.1002/psp4.12875>
* Supplement (Monolix project files for each estimation step, parameter Tables S1-S3): <https://doi.org/10.1002/psp4.12875>

This is a semi-mechanistic pharmacokinetic-pharmacodynamic (a quantitative
systems-pharmacology-style) model of the antimalarial DSM265, a *Plasmodium
falciparum* dihydroorotate-dehydrogenase (DHODH) inhibitor developed as a
chemoprophylactic candidate. The model couples a two-compartment PK model for
DSM265 to a two-stage parasite-dynamics model (liver / hepatic-schizont stage
and blood stage), each growing exponentially and killed by a turnover-lagged
Emax / EC50 Hill term driven by the drug concentration. Merozoites move from
liver to blood through a sigmoidal maturation window centred at T50 ~ 6 days.

# Population

Data were pooled from 15 studies comprising 190 participants (Cherkaoui-Rbati
2023, Table 1): one first-in-human single-ascending-dose study (55 subjects,
25-1200 mg), two induced-blood-stage malaria (IBSM) studies (15), a phase IIa
study of uncomplicated *P. falciparum* malaria patients in Peru (24), two
sporozoite human-challenge (spz HuCh) studies (39, including 10 placebo), and 57
placebo subjects from nine further published sporozoite-challenge studies
(Coffeng et al. 2017). The chemoprophylaxis arms of the spz HuCh studies dosed
400 mg DSM265 at 1, 3, or 7 days before intravenous inoculation of 3200
sporozoites.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("CherkaouiRbati_2023_dsm265")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in
`inst/modeldb/specificDrugs/CherkaouiRbati_2023_dsm265.R`. The structural
equations and fixed values were confirmed against the deposited Monolix project
files (supplement, Step 1-5 projects). The table below collects the key origins.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL)           | 0.476 L/h | Table 2 |
| `lvc` (Vc)           | 8.58 L | Table 2 |
| `lq` (Q1)            | 37.1 L/h | Table 2 |
| `lvp` (Vp,1)         | 57.3 L | Table 2 |
| `ld1` (Tk,0)         | 2.85 h | Table 2 |
| `ltlag` (Tlag,1)     | 0.147 h | Table 2 |
| `lfrel` (F)          | 1 (fixed) | Table 2 |
| `e_dose_frel`        | -0.102 | Table 2 (beta_F,Dose; transform log(dose/400)) |
| `lkgrow_liver` (GR_L) | 0.0716 1/h (fixed) | Table 3 |
| `lkgrow_blood` (GR_B) | 0.0624 1/h (fixed) | Table 3 / Table S2 |
| `lemax_liver`/`lemax_blood` (Emax) | 0.205 1/h (fixed) | Table 3 / Table S1 |
| `lec50_liver` (EC50,L) | 1.7 ug/mL | Table 3 (estimated) |
| `lec50_blood` (EC50,B) | 1.25 ug/mL (fixed) | Table 3 / Table S1 |
| `lhill_liver`/`lhill_blood` (h) | 10 (fixed) | Table 3 / Table S1 |
| `lkin` (k_in)        | 0.0771 1/h (fixed) | Table 3 / Table S1 |
| `lklb` (k_LB)        | 6 1/h (fixed) | Table 3 |
| `lt50` (T50)         | 144 h (fixed) | Table 3 |
| `lsigtliver` (sigma_LB) | 0.1 h (fixed) | Table 3 |
| `lfinc` (F_inc)      | 0.00119 (fixed) | Table 3 / Table S3 |
| `lrhogr` / `lrhoec50` (correlation scalings) | 1 (fixed) | Table 3 (rho = 1) |
| `propSd`             | 0.179 | Table 2 (proportional error) |
| `addSd`              | 2.66 ln(p/mL) | Table 3 (additive error) |
| `d/dt(parasite_liver)`, `d/dt(parasite_blood)` | n/a | Equations 1.1, 1.2 |
| `d/dt(kill_liver)`, `d/dt(kill_blood)` | n/a | Equations 2.1, 2.2 |
| `trlb` (merozoite release) | n/a | Equation 3 |

# Model conventions and the time origin

The model uses **absolute time `t` with sporozoite infection at `t = 0`**, which
drives the merozoite-release window `trlb = expit((t - T50) / sigma_LB)`
(T50 = 144 h). Chemoprophylactic drug doses are therefore administered at
**negative times** relative to infection (e.g. -24 h for a Day -1 dose),
matching the convention of the source data. The sporozoite inoculum is
administered as a dose into `parasite_liver`, with bioavailability `F_inc`
applied, so the initial liver burden is `F_inc * inoculum` (for 3200 sporozoites
that is ~4 infected hepatocytes, Cherkaoui-Rbati 2023 Results). Mosquito-bite
dosing should supply the inoculum in sporozoites (bites x 640 spz/bite).

# PK validation (DSM265 disposition)

A single 400 mg oral dose with no infection reproduces the long terminal
half-life (~96-100 h, Cherkaoui-Rbati 2023; "a long elimination half-life in
humans of around 96 h").

```{r pk-sim}
rxode2::rxSetSeed(1024)
set.seed(1024)

mod <- readModelDb("CherkaouiRbati_2023_dsm265")

# The model has two residual-error endpoints (Cc and lparasitemia_blood), so
# rxode2 requires every observation row to name one of them in `cmt`; an
# observation row on an ODE state such as "central" is rejected. Both
# observables, and every state, are still returned as columns on every row.
n_pk <- 150L
pk_obs_times <- c(seq(0, 48, by = 1), seq(52, 480, by = 4))
pk_dose <- data.frame(
  id = seq_len(n_pk), time = 0, evid = 1L, amt = 400, cmt = "central"
)
pk_obs <- tidyr::crossing(
  id = seq_len(n_pk), time = pk_obs_times
) |>
  dplyr::mutate(evid = 0L, amt = NA_real_, cmt = "Cc")
pk_events <- dplyr::bind_rows(pk_dose, pk_obs) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  dplyr::mutate(treatment = "400 mg")

pk_sim <- rxode2::rxSolve(mod, events = pk_events, keep = "treatment") |>
  as.data.frame()
```

```{r fig-pk}
pk_sim |>
  dplyr::filter(!is.na(Cc), time > 0) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "DSM265 concentration (ug/mL)",
    title = "DSM265 PK after a single 400 mg oral dose",
    caption = "Two-compartment model, Cherkaoui-Rbati 2023 Table 2."
  )
```

```{r pknca}
sim_nca <- pk_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

# Defensive time-zero rows (extravascular pre-dose Cc = 0).
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |>
    dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "ug/mL", timeu = "h")

dose_df <- pk_events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id, doseu = "mg")

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_summary <- as.data.frame(nca_res$result) |>
  dplyr::group_by(PPTESTCD) |>
  dplyr::summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop")
nca_summary |>
  dplyr::rename("NCA parameter" = PPTESTCD, "Median (simulated)" = median) |>
  knitr::kable(
    digits = 3,
    caption = "Simulated single-dose NCA (400 mg DSM265)."
  )
```

```{r pknca-check}
hl <- nca_res$result |>
  dplyr::filter(PPTESTCD == "half.life") |>
  dplyr::pull(PPORRES)

# Typical-value terminal half-life from the Table 2 estimates (the slower
# eigenvalue of the two-compartment system). This is deterministic, so it is
# held tightly against the "around 96 h" the paper quotes.
pk_ini <- rxode2::rxode(mod)$theta
k <- exp(pk_ini[["lcl"]]) / exp(pk_ini[["lvc"]])
k12 <- exp(pk_ini[["lq"]]) / exp(pk_ini[["lvc"]])
k21 <- exp(pk_ini[["lq"]]) / exp(pk_ini[["lvp"]])
s <- k + k12 + k21
beta <- (s - sqrt(s^2 - 4 * k * k21)) / 2
t_half_typical <- log(2) / beta
t_half_typical

# The cohort median runs longer than the typical value: the very large
# variance on Vc (SD 1.06 on the log scale) skews (Vc + Vp) / CL upwards. The
# cohort check is therefore an envelope only.
median(hl, na.rm = TRUE)
stopifnot(
  abs(t_half_typical - 96) < 3,
  median(hl, na.rm = TRUE) > 80,
  median(hl, na.rm = TRUE) < 160
)
```

# Chemoprophylaxis: parasite dynamics

The next simulation reproduces the two-stage parasite dynamics for a single
400 mg dose given 7 days before infection: liver parasites grow and then burst
near Day 6, seeding the blood stage, which either grows (treatment failure) or is
suppressed (protection).

```{r pd-single}
pd_obs_times <- seq(-168, 28 * 24, by = 6)
one_dose <- data.frame(id = 1, time = -168, evid = 1L, amt = 400, cmt = "central")
one_spz <- data.frame(id = 1, time = 0, evid = 1L, amt = 3200, cmt = "parasite_liver")
one_obs <- data.frame(
  id = 1, time = pd_obs_times, evid = 0L, amt = NA_real_, cmt = "lparasitemia_blood"
)
one_events <- dplyr::bind_rows(one_dose, one_spz, one_obs) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

one_mod <- mod |> rxode2::ini(
  etalcl = 0, etalvc = 0, etalvp = 0, etald1 = 0, etaltlag = 0,
  etalfrel = 0, etalfinc = 0, etalrhogr = 0, etalrhoec50 = 0
)
one_sim <- rxode2::rxSolve(one_mod, events = one_events) |>
  as.data.frame()
```

```{r fig-pd-single}
one_sim |>
  dplyr::transmute(
    time_day = time / 24,
    `Liver parasites (count)` = pmax(parasite_liver, 1e-3),
    `Blood parasitemia (p/mL)` = pmax(parasite_blood / 5000, 1e-3)
  ) |>
  tidyr::pivot_longer(-time_day) |>
  ggplot(aes(time_day, value, colour = name)) +
  geom_line() +
  scale_y_log10() +
  labs(
    x = "Time since infection (days)", y = "Parasites",
    colour = NULL,
    title = "Two-stage parasite dynamics (typical individual, 400 mg at Day -7)",
    caption = "Liver burst ~Day 6 seeds the blood stage (Cherkaoui-Rbati 2023 Fig 2a)."
  ) +
  theme(legend.position = "bottom")
```

# Chemoprophylaxis: Day-28 protection success rates (Figure 3)

The paper's central validation is the Day-28 protection success rate for 400 mg
DSM265 given at Day -1, -3, or -7 before infection. "Success" is being
malaria-free at Day 28, i.e. blood parasitemia below the 20 p/mL PCR limit of
quantification.

```{r success-cohort}
n_arm <- 150L
arms <- tibble::tribble(
  ~arm,      ~dose_time,
  "Day -1",  -24,
  "Day -3",  -72,
  "Day -7",  -168
)

make_arm <- function(arm_label, dose_time, id_offset) {
  ids <- id_offset + seq_len(n_arm)
  drug <- data.frame(id = ids, time = dose_time, evid = 1L, amt = 400, cmt = "central")
  spz <- data.frame(id = ids, time = 0, evid = 1L, amt = 3200, cmt = "parasite_liver")
  # One observation at Day 28 (the landmark) plus a coarse grid for stability.
  obs <- tidyr::crossing(id = ids, time = c(seq(0, 28 * 24, by = 24))) |>
    dplyr::mutate(evid = 0L, amt = NA_real_, cmt = "lparasitemia_blood")
  dplyr::bind_rows(drug, spz, obs) |>
    dplyr::mutate(arm = arm_label)
}

success_events <- purrr::pmap_dfr(
  list(arms$arm, arms$dose_time, (seq_len(nrow(arms)) - 1L) * n_arm),
  make_arm
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))
stopifnot(!anyDuplicated(unique(success_events[, c("id", "time", "evid")])))

success_sim <- rxode2::rxSolve(mod, events = success_events, keep = "arm") |>
  as.data.frame()
```

```{r success-rate}
success_rate <- success_sim |>
  dplyr::filter(abs(time - 28 * 24) < 1e-6) |>
  dplyr::mutate(parasitemia = parasite_blood / 5000) |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    n = dplyr::n(),
    success_pct = 100 * mean(parasitemia < 20, na.rm = TRUE),
    .groups = "drop"
  )

published <- tibble::tribble(
  ~arm,     ~obs_pct, ~pred_pct,
  "Day -1", 100,      60,
  "Day -3", 33,       33,
  "Day -7", 39,       11
)

success_rate |>
  dplyr::left_join(published, by = "arm") |>
  dplyr::rename(
    "Arm" = arm,
    "N" = n,
    "Simulated success (%)" = success_pct,
    "Observed (%) [Fig 3]" = obs_pct,
    "Published model (%) [Fig 3]" = pred_pct
  ) |>
  knitr::kable(digits = 1, caption = "Day-28 protection success rate vs. Cherkaoui-Rbati 2023 Figure 3.")
```

```{r success-check}
sr <- stats::setNames(success_rate$success_pct, success_rate$arm)
# The paper's estimated-model Day-28 success rates rise from ~11% (Day -7) to
# ~60% (Day -1). Assert each arm lands inside a generous envelope around the
# published 90% CIs, and that earlier-than-infection dosing (Day -7) protects
# far LESS than near-infection dosing (Day -1) -- a ~50-point gap the paper
# reports, which is robust to the drawn cohort. Do NOT tighten these to a single
# run's values (pattern 12 of known-vignette-failure-patterns.md).
stopifnot(
  sr[["Day -7"]] >= 0, sr[["Day -7"]] <= 45,
  sr[["Day -1"]] >= 25, sr[["Day -1"]] <= 100,
  sr[["Day -1"]] > sr[["Day -7"]]
)
```

The simulated success rates track the published model predictions (about 60%,
33% and 11% for Day -1, -3 and -7; Figure 3, blue) to within a few percentage
points. That is within the binomial noise of a 150-subject arm, and well inside
the published 90% intervals. Like the published model, the simulation
under-predicts the observed rate at Day -1 (5 of 5 protected) and at Day -7
(7 of 18). The authors attribute the Day -1 gap to the continuous (rather than
stochastic) description of the few parasites present early in infection. The
broader pattern also holds: protection improves
as the first dose moves closer to infection, because the later dose keeps drug
concentrations above the protective threshold through the Day-6 liver burst.
The paper notes that, even at Day -1, protection is incomplete because the MIC
is the concentration at which killing equals growth (not a parasiticidal
concentration) and because of interindividual PK/PD variability.

# Assumptions and deviations

- The model uses absolute time with infection fixed at `t = 0`; prophylactic
  doses are given at negative times. This matches the source data convention and
  keeps the merozoite-release window `T50 = 144 h` interpretable as 6 days after
  infection.
- The liver-stage Emax, Hill coefficient, and turnover rate were fixed to the
  blood-stage values, and the liver/blood growth rates and EC50 were taken as
  100% correlated within an individual, exactly as reported (Cherkaoui-Rbati
  2023 Table 3; these liver parameters were not separately identifiable from the
  available data). Only the liver-stage EC50 and its variability were estimated
  in the final model.
- Final parameter values were taken from Table 2 (PK) and Table 3 (PD). The
  deposited Monolix project files (supplement) carry the same fixed values to
  more digits and were used to confirm the ODE structure, the sigmoidal release
  term, and the dosing of the sporozoite inoculum into the liver pool; the
  estimated typicals in those project files are initial values, so the Table
  2 / Table 3 point estimates are authoritative.
- The liver-stage EC50 is 1.7 ug/mL in Table 3 but "1.6 ug/ml" in the Results
  text, which also quotes a liver-stage MIC of 1.5 ug/mL. The Table 3 value is
  used. By Equation 5, MIC_L = EC50_L * (GR_L / (Emax_L - GR_L))^(1/h_L) = 0.940 *
  EC50_L. That gives a typical MIC_L of 1.60 ug/mL with the Table 3 EC50 and
  1.50 ug/mL with the text's, so the text's two numbers agree with each other
  and the table differs from both. The two values differ by 6%.
- The PK model's individual parameters are, in the source, carried into the PD
  layer as subject-level regressors. Here they are represented by the Table 2
  population PK model with its interindividual variability, so the packaged
  model is self-contained and can simulate PK and PD jointly.
- The sporozoite inoculum is dosed into `parasite_liver` in sporozoites, with
  `F_inc` as its bioavailability. Mosquito-bite dosing (640 sporozoites per
  bite) is supported by supplying the inoculum already converted to sporozoites.
- The additive residual error (2.66 on the natural-log scale of parasitemia) is
  large because most parasitemia samples were below the limit of quantification
  and were handled by the M3 likelihood method; it is reported faithfully here.
- No correction or erratum notice for this article was found as of 2026-10-11.
```
