---
title: "Bamlanivimab + etesevimab PK and SARS-CoV-2 viral dynamics (Chigutsa 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bamlanivimab + etesevimab PK and SARS-CoV-2 viral dynamics (Chigutsa 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Chigutsa_2021_bamlanivimab_etesevimab"))$reference`
* Article (open access): <https://doi.org/10.1002/cpt.2420>
* Supporting Information: Table S1 (demographics), Table S2 (PK
  parameters) and the NONMEM control stream of the viral-dynamic model.

Bamlanivimab and etesevimab are neutralizing IgG1 antibodies against the
receptor-binding domain of the SARS-CoV-2 spike protein. Chigutsa 2021
analysed them sequentially. First, each antibody got its own
two-compartment PK model. Then the individual post hoc PK predictions
drove a target-cell-limited viral-dynamic model of nasopharyngeal viral
load. In that model both antibodies raise the virus elimination rate.

The package ships all three parts as one model,
`Chigutsa_2021_bamlanivimab_etesevimab`. This matches the published
viral-dynamic control stream, whose `$DES` block carries both PK
systems next to the viral states. The two PK systems share no
parameters, so the model can also be used for either antibody's PK
alone.

```{r model}
mod <- rxode2::rxode2(readModelDb("Chigutsa_2021_bamlanivimab_etesevimab"))
mod_typ <- rxode2::zeroRe(mod)
```

# Population

The PK analysis pooled BLAZE-1 (NCT04427501) and BLAZE-4 (NCT04634409),
two phase II dose-ranging, placebo-controlled trials in outpatients with
mild to moderate COVID-19. It also included 18 hospitalized patients
with severe COVID-19 (J2W-MC-PYAA) and 20 healthy participants
(J2Z-MC-PGAA). The data were 5915 bamlanivimab concentrations from 1899
participants and 4961 etesevimab concentrations from 1498 participants.
The viral-dynamic analysis used 17,805 viral-load measurements from 2970
BLAZE-1/-4 participants, 1047 of whom received placebo. Single IV doses
ranged from 175 to 7000 mg bamlanivimab, alone or with 350 to 2800 mg
etesevimab.

In the viral-load population (Table S1), median body weight was 87.8 kg
(range 41.4-220) and median age 50 years (12-94). 52.4% were female and
85.9% White, 7.2% Black and 3.9% Asian. 20.2% had mild hepatic impairment.
Median baseline viral load was 5.30 log10 units, where one unit is
(40 - Ct)/log2(10).

# Source trace

| Quantity | Value | Source |
|---|---|---|
| Bamlanivimab CL, Q, V1, V2 (70 kg) | 0.231 L/d, 0.281 L/d, 2.68 L, 2.68 L | Table S2 |
| Etesevimab CL, Q, V1, V2 (70 kg) | 0.111 L/d, 0.308 L/d, 2.45 L, 2.18 L | Table S2 |
| Weight exponent on CL and Q / on V1 and V2 | 0.81 fixed / 1 fixed | Table S2 footnotes a, b |
| IIV CL, V1 (bamlanivimab; etesevimab) | 25.1%, 28.9%; 28.4%, 30.7% | Table S2 |
| Proportional residual error (bam; ete) | 19.0%; 18.3% | Table S2 |
| Target-cell pool | 4 x 10^8 fixed (stream: 400 in units of 1e6) | Table 1; stream THETA(1) |
| Log10 viral load at symptom onset | 7.60 | Table 1 |
| beta, PV, CV, DI | 5.23e-7, 0.0844 /d, 1.42 /d, 0.290 /d | Table 1 |
| Emax, EC50 (bamlanivimab) | 0.462, 0.467 ug/mL | Table 1 |
| Etesevimab EC50 = 3 x bamlanivimab EC50 | fixed | Results "Viral dynamic modeling"; stream `$DES` |
| IIV beta, PV, CV and their correlations | 2526%, 47.2%, 85.4%; 0.939, 0.344, 0.359 | Table 1 |
| IIV viral load at onset, DI, Emax, EC50 | 15% fixed each | Table 1; stream `$OMEGA` 0.0225 FIX |
| Additive residual error (log10 viral load) | 0.939 | Table 1; stream `$ERROR` |
| Two-compartment PK ODEs | -- | Methods; stream `$DES` A(1), A(2), A(6), A(7) |
| Target-cell-limited viral ODEs | -- | Methods equations; stream `$DES` A(3)-A(5) |
| CV(t) = CV + Emax C1/(EC50 + C1) + Emax C2/(3 EC50 + C2) | -- | Table 1 footnote c; stream `TEFF`, `CBTEFF` |
| log10 viral load = log10(virus x 1e6), floored at 1e-5 | -- | stream `$ERROR` |

# Typical-value viral dynamics

## Replicating Table 2

Table 2 of the paper gives the model-predicted reduction in viral load
relative to placebo after 700 mg bamlanivimab plus 1400 mg etesevimab
given 1, 4 or 10 days after symptom onset. The model clock starts at
symptom onset. The time-weighted-average windows "days 0 to 7" and
"days 0 to 11" start on the **day of treatment**. (A window starting at
symptom onset does not reproduce the table.) The infusion duration is
not reported, so a 1-hour infusion is used here.

```{r helpers}
# One subject with observations on a fine grid; dvid = 3 nominates the
# log10_viral_load endpoint, and every endpoint comes back as a column.
make_events <- function(t_dose = NA, bam = 0, ete = 0, wt = 70, id = 1,
                        times = seq(0, 30, by = 0.01)) {
  obs <- data.frame(
    id = id, time = times, amt = 0, rate = 0, evid = 0L,
    cmt = "virus", dvid = 3L, WT = wt
  )
  if (is.na(t_dose)) {
    return(obs)
  }
  dose <- data.frame(
    id = id, time = t_dose, amt = c(bam, ete), rate = c(bam, ete) * 24,
    evid = 1L, cmt = c("central", "central_ete"), dvid = NA_integer_, WT = wt
  )
  dose <- dose[dose$amt > 0, ]
  ev <- rbind(dose, obs)
  ev[order(ev$id, ev$time, -ev$evid), ]
}

solve_typ <- function(ev) {
  as.data.frame(rxode2::rxSolve(mod_typ, ev, returnType = "data.frame"))
}

# Trapezoidal time-weighted average of y over [lo, hi]
twa <- function(t, y, lo, hi) {
  keep <- t >= lo - 1e-9 & t <= hi + 1e-9
  tt <- t[keep]
  yy <- y[keep]
  sum(diff(tt) * (head(yy, -1) + tail(yy, -1)) / 2) / (hi - lo)
}
```

```{r table2}
placebo <- solve_typ(make_events())

table2 <- lapply(c(1, 4, 10), function(td) {
  trt <- solve_typ(make_events(t_dose = td, bam = 700, ete = 1400))
  data.frame(
    day = td,
    twa7 = twa(placebo$time, placebo$log10_viral_load, td, td + 7) -
      twa(trt$time, trt$log10_viral_load, td, td + 7),
    twa11 = twa(placebo$time, placebo$log10_viral_load, td, td + 11) -
      twa(trt$time, trt$log10_viral_load, td, td + 11),
    maxred = max(placebo$log10_viral_load - trt$log10_viral_load)
  )
}) |>
  bind_rows()

published <- data.frame(
  day = c(1, 4, 10),
  twa7_pub = c(1.15, 0.70, 0.25),
  twa11_pub = c(0.99, 0.55, 0.24),
  maxred_pub = c(1.82, 1.15, 0.28)
)

t2 <- left_join(table2, published, by = "day")
t2 |>
  mutate(across(where(is.double), \(x) round(x, 2))) |>
  select(day, twa7_pub, twa7, twa11_pub, twa11, maxred_pub, maxred) |>
  rename(
    "Treatment day (from symptom onset)" = day,
    "TWA days 0-7, paper" = twa7_pub,
    "TWA days 0-7, model" = twa7,
    "TWA days 0-11, paper" = twa11_pub,
    "TWA days 0-11, model" = twa11,
    "Max reduction, paper" = maxred_pub,
    "Max reduction, model" = maxred
  ) |>
  knitr::kable(caption = "Replicates Table 2 of Chigutsa 2021: log10 reduction relative to placebo.")
```

All nine cells are deterministic typical-value solves, so the comparison
is tight. The paper prints two decimals. The largest gap (0.017 log10,
TWA days 0-7 after treatment on day 1) is within rounding plus the
unreported infusion duration.

```{r table2-check}
stopifnot(
  all(abs(t2$twa7 - t2$twa7_pub) < 0.03),
  all(abs(t2$twa11 - t2$twa11_pub) < 0.03),
  all(abs(t2$maxred - t2$maxred_pub) < 0.03)
)
```

The in vivo IC90 values in the Abstract follow directly from the EC50s
with Hill = 1 (IC90 = 9 x EC50). The same holds for the 25.2 ug/mL used
in the Figure S2 sensitivity analysis, where etesevimab is assumed 6-fold
less potent than bamlanivimab.

```{r ic90}
ec50 <- exp(mod$theta[["lec50"]])
ratio <- mod$theta[["ec50_ratio_ete"]]
ic90 <- c(bamlanivimab = 9 * ec50, etesevimab = 9 * ec50 * ratio, etesevimab_6x = 9 * ec50 * 6)
round(ic90, 1)
stopifnot(
  abs(ic90[["bamlanivimab"]] - 4.2) < 0.05,
  abs(ic90[["etesevimab"]] - 12.6) < 0.05,
  abs(ic90[["etesevimab_6x"]] - 25.2) < 0.05
)
```

## Replicating Figure 4: timing and monotherapy vs combination

```{r figure4}
fig4 <- bind_rows(
  lapply(c(1, 4, 10), function(td) {
    bind_rows(
      solve_typ(make_events(t_dose = td, bam = 700)) |>
        mutate(arm = "Bamlanivimab 700 mg"),
      solve_typ(make_events(t_dose = td, bam = 700, ete = 1400)) |>
        mutate(arm = "Bamlanivimab 700 mg + etesevimab 1400 mg")
    ) |>
      mutate(t_dose = paste("Treated on day", td))
  })
) |>
  left_join(
    placebo |> select(time, placebo = log10_viral_load),
    by = "time"
  ) |>
  mutate(difference = log10_viral_load - placebo)

placebo_panel <- tidyr::crossing(
  placebo |> select(time, log10_viral_load),
  t_dose = unique(fig4$t_dose)
) |>
  mutate(arm = "Placebo")

ggplot(bind_rows(fig4, placebo_panel), aes(time, log10_viral_load, colour = arm)) +
  geom_line() +
  facet_wrap(~t_dose) +
  coord_cartesian(xlim = c(0, 20)) +
  labs(
    x = "Days since onset of symptoms", y = "log10 viral load", colour = NULL,
    caption = "Replicates Figure 4 (top) of Chigutsa 2021."
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

ggplot(fig4, aes(time, difference, colour = arm)) +
  geom_line() +
  facet_wrap(~t_dose) +
  coord_cartesian(xlim = c(0, 20)) +
  labs(
    x = "Days since onset of symptoms", y = "Difference from placebo (log10)",
    colour = NULL, caption = "Replicates Figure 4 (bottom) of Chigutsa 2021."
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

Adding etesevimab gives a further reduction beyond bamlanivimab alone.
This matches the paper's reading that bamlanivimab exposure already sits
near its own Emax. The etesevimab term adds a second Emax.

```{r figure4-check}
day7 <- fig4 |>
  filter(abs(time - 7) < 1e-6, t_dose == "Treated on day 4") |>
  select(arm, log10_viral_load)
day7
stopifnot(
  day7$log10_viral_load[day7$arm == "Bamlanivimab 700 mg + etesevimab 1400 mg"] <
    day7$log10_viral_load[day7$arm == "Bamlanivimab 700 mg"] - 0.3
)
```

## Replicating Figure 3: body weight

Figure 3 compares a 40 kg and a 220 kg patient given 700/1400 mg. Here
the dose is on day 4 after symptom onset (the trials' median). Serum
concentrations differ several-fold, but both stay far above the IC90 for
weeks. The viral-load curves overlap.

```{r figure3}
fig3 <- bind_rows(
  solve_typ(make_events(t_dose = 4, bam = 700, ete = 1400, wt = 40, times = seq(0, 40, by = 0.05))) |>
    mutate(weight = "40 kg"),
  solve_typ(make_events(t_dose = 4, bam = 700, ete = 1400, wt = 220, times = seq(0, 40, by = 0.05))) |>
    mutate(weight = "220 kg")
)

fig3 |>
  select(time, weight, `log10 viral load` = log10_viral_load, `Bamlanivimab (ug/mL)` = Cc, `Etesevimab (ug/mL)` = Cc_ete) |>
  pivot_longer(-c(time, weight)) |>
  ggplot(aes(time, value, colour = weight, linetype = weight)) +
  geom_line() +
  facet_wrap(~name, ncol = 1, scales = "free_y") +
  labs(x = "Days since onset of symptoms", y = NULL, colour = "Body weight", linetype = "Body weight",
       caption = "Replicates Figure 3 of Chigutsa 2021 (dose on day 4).") +
  theme_bw()

vl_gap <- fig3 |>
  select(time, weight, log10_viral_load) |>
  pivot_wider(names_from = weight, values_from = log10_viral_load)
max_gap <- max(abs(vl_gap$`40 kg` - vl_gap$`220 kg`))
c28 <- fig3 |> filter(abs(time - 32) < 1e-6, weight == "220 kg")
c(max_vl_gap = max_gap, bam_day28_220kg = c28$Cc, ete_day28_220kg = c28$Cc_ete)
stopifnot(
  max_gap < 0.05,
  c28$Cc > ic90[["bamlanivimab"]],
  c28$Cc_ete > ic90[["etesevimab"]]
)
```

# Stochastic PK: virtual cohort

## Virtual cohort

The cohort is 200 virtual patients. Body weight is drawn log-normally
around the viral-load population's median of 87.8 kg, and draws outside
the observed 41.4-220 kg range are rejected and redrawn. Each patient
gets 700 mg bamlanivimab plus 1400 mg etesevimab on day 4 after symptom
onset. Serum is sampled on the trials' schedule, extended to a dense
grid for NCA.

```{r cohort}
rxode2::rxSetSeed(20211)
n_sub <- 200
draw_wt <- function(n) {
  out <- numeric(0)
  while (length(out) < n) {
    w <- exp(rnorm(n, log(87.8), 0.25))
    out <- c(out, w[w >= 41.4 & w <= 220])
  }
  out[seq_len(n)]
}
wts <- draw_wt(n_sub)
summary(wts)

t_dose <- 4
obs_times <- t_dose + c(0, 1 / 24, 0.25, 0.5, 1, 2, 4, 7, 10, 14, 21, 28, 35, 42, 56, 70, 84)
cohort_ev <- lapply(seq_len(n_sub), function(i) {
  make_events(t_dose = t_dose, bam = 700, ete = 1400, wt = wts[i], id = i, times = c(0, obs_times))
}) |>
  bind_rows()

sim <- rxode2::rxSolve(mod, cohort_ev, returnType = "data.frame") |>
  as.data.frame() |>
  mutate(tad = time - t_dose, treatment = "700 mg bamlanivimab + 1400 mg etesevimab")
```

## Replicating Figure 5 (right): PK profiles against the IC90

```{r figure5}
pk_long <- sim |>
  filter(tad >= 0) |>
  select(id, tad, Bamlanivimab = Cc, Etesevimab = Cc_ete) |>
  pivot_longer(c(Bamlanivimab, Etesevimab), names_to = "drug", values_to = "conc")

pk_sum <- pk_long |>
  group_by(drug, tad) |>
  summarise(
    median = median(conc), lo = quantile(conc, 0.05), hi = quantile(conc, 0.95),
    .groups = "drop"
  )
ic90_band <- data.frame(
  drug = c("Bamlanivimab", "Etesevimab"),
  lo = c(3.2, 9.7), hi = c(4.3, 12.8)
)

ggplot(pk_sum, aes(tad, median)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.3) +
  geom_line() +
  geom_rect(
    data = ic90_band, inherit.aes = FALSE,
    aes(xmin = -Inf, xmax = Inf, ymin = lo, ymax = hi), alpha = 0.3, fill = "red"
  ) +
  facet_wrap(~drug) +
  scale_y_log10() +
  labs(
    x = "Days after dose", y = "Serum concentration (ug/mL)",
    caption = "Replicates Figure 5 (right) of Chigutsa 2021: 90% PI; red band = 95% CI of the IC90."
  ) +
  theme_bw()
```

Figure 5 (left) reports that at least 90% of patients stay above the
IC90 for 28 days after the authorized doses. The simulated day-28
proportions are shown below. The gate is on the day-28 median, which is
far above the IC90; a threshold-crossing percentage is not a stable
gate.

```{r pta}
day28 <- pk_long |> filter(abs(tad - 28) < 1e-6)
day28 |>
  group_by(drug) |>
  summarise(
    median_conc = median(conc),
    pct_above_ic90 = 100 * mean(conc > ifelse(drug == "Bamlanivimab", ic90[["bamlanivimab"]], ic90[["etesevimab"]])),
    .groups = "drop"
  ) |>
  rename("Drug" = drug, "Median day-28 conc (ug/mL)" = median_conc, "% above IC90" = pct_above_ic90) |>
  knitr::kable(digits = 1)

stopifnot(
  median(day28$conc[day28$drug == "Bamlanivimab"]) > 3 * ic90[["bamlanivimab"]],
  median(day28$conc[day28$drug == "Etesevimab"]) > 3 * ic90[["etesevimab"]]
)
```

# PKNCA validation

The paper reports one NCA-type quantity for each antibody: the terminal
half-life, 20.9 days for bamlanivimab and 32.6 days for etesevimab
(Results, "PK modeling"). The paper does not say which summary statistic
this is. Here the simulated median is compared against it. The model
predictions are used without residual error.

```{r pknca}
nca_one <- function(col, label) {
  conc <- sim |>
    filter(tad >= 0, !is.na(.data[[col]])) |>
    transmute(id, time = tad, conc = .data[[col]], treatment = label) |>
    distinct(id, time, .keep_all = TRUE)
  dose <- conc |>
    distinct(id, treatment) |>
    mutate(time = 0, amt = ifelse(label == "Bamlanivimab", 700, 1400))
  intervals <- data.frame(
    start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
    auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE
  )
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(conc, conc ~ time | treatment + id),
    PKNCA::PKNCAdose(dose, amt ~ time | treatment + id),
    intervals = intervals
  ))
  as.data.frame(res$result)
}

nca_res <- bind_rows(
  nca_one("Cc", "Bamlanivimab"),
  nca_one("Cc_ete", "Etesevimab")
)

reference <- data.frame(
  treatment = c("Bamlanivimab", "Etesevimab"),
  half.life = c(20.9, 32.6)
)

cmp <- ncaComparisonTable(
  nca_res, reference,
  by = "treatment", params = "half.life",
  units = c(half.life = "day")
)
knitr::kable(cmp)
```

```{r pknca-check}
hl <- nca_res |>
  filter(PPTESTCD == "half.life") |>
  group_by(treatment) |>
  summarise(median_hl = median(PPORRES), .groups = "drop") |>
  left_join(reference, by = "treatment") |>
  mutate(pct_diff = 100 * (median_hl - half.life) / half.life)
hl
stopifnot(all(abs(hl$pct_diff) < 10))
```

As a deterministic cross-check, the typical-value terminal half-life
at each drug's median body weight (85.7 kg bamlanivimab PK population,
87.8 kg etesevimab) is:

```{r halflife-typical}
typ_hl <- function(cl, q, v1, v2, wt) {
  cl <- cl * (wt / 70)^0.81
  q <- q * (wt / 70)^0.81
  v1 <- v1 * wt / 70
  v2 <- v2 * wt / 70
  k10 <- cl / v1
  k12 <- q / v1
  k21 <- q / v2
  s <- k10 + k12 + k21
  log(2) / ((s - sqrt(s^2 - 4 * k10 * k21)) / 2)
}
th <- mod$theta
hl_typ <- c(
  bamlanivimab = typ_hl(exp(th[["lcl"]]), exp(th[["lq"]]), exp(th[["lvc"]]), exp(th[["lvp"]]), 85.7),
  etesevimab = typ_hl(exp(th[["lcl_ete"]]), exp(th[["lq_ete"]]), exp(th[["lvc_ete"]]), exp(th[["lvp_ete"]]), 87.8)
)
round(hl_typ, 1)
stopifnot(abs(hl_typ[["bamlanivimab"]] - 20.9) < 0.5, abs(hl_typ[["etesevimab"]] - 32.6) < 0.5)
```

# Assumptions and deviations

* **IIV scale.** Tables 1 and S2 report IIV as %CV. The beta entry of
  2526% is only consistent with CV = sqrt(exp(omega^2) - 1). The "15%
  Fixed" entries are `OMEGA 0.0225 FIX` in the control stream, which
  gives 15.1% on the same scale. So omega^2 = log(1 + CV^2) is used for
  every IIV term, including the PK terms of Table S2. The beta-PV-CV
  covariance block is rebuilt from the reported CVs and correlations.
* **No PK correlations.** Table S2 reports no CL-V1 correlation, so the
  PK etas are independent. IIV on Q and V2 was not estimated.
* **Stream scale for the viral states.** The control stream holds the
  virus and both cell pools in units of 1e6. The observation is
  `log10(virus * 1e6)`, and the target-cell THETA is 400 FIX for the
  4 x 10^8 cells in Table 1. The Table 1 values of beta, PV, CV and DI
  are the stream's THETAs and are used on that scale. The Table 1 unit
  label for beta, ((copies/mL)^-1 day^-1), refers to the linear viral
  load before the 1e6 scaling. Table 2 is reproduced within rounding,
  which confirms this reading. Viral load at symptom onset (7.60 log10)
  enters the model as 10^(7.60 - 6).
* **Target-cell IIV.** The stream fixes the target-cell IIV to 0
  (`$OMEGA 0 FIX`). The model therefore has no eta on the target-cell
  pool rather than a zero-variance eta.
* **Emax is an additive rate.** Table 1 labels Emax "fractional increase
  in CV". Footnote c and the stream's `$DES` both add
  `Emax * C / (EC50 + C)` to CV as a rate (per day). The equation is
  used here.
* **Etesevimab potency.** The etesevimab EC50 is fixed at three times the
  bamlanivimab EC50, as in the paper. It is exposed as
  `ec50_ratio_ete = fixed(3)` so the Figure S2 sensitivity analysis
  (6-fold) can be run by overriding it.
* **Concentration floor.** The stream floors the serum concentration at
  1e-6 ug/mL when it is not positive. Its effect on the Emax terms is
  negligible, so it is omitted.
* **Censoring.** The analysis fitted viral loads below 0.05 log10 units
  with Beal's M4 method, truncated at 0. This belongs to estimation, not
  to the structural model. To refit, supply the censoring columns in the
  data.
* **Infusion duration.** Not reported. Simulations here use a 1-hour
  infusion. It barely affects any of the comparisons above.
* **Viral-load units.** The dependent variable is (40 - Ct)/log2(10), a
  Ct-derived log10 scale relative to a cycle threshold of 40. It is not
  a calibrated copies/mL value.
* **Weight distribution.** Table S1 gives the weight median and range
  but no spread. The log-normal SD of 0.25 used for the virtual cohort is
  the maintainers' assumption.
