---
title: "Efpeglenatide (Choi 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Efpeglenatide (Choi 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
ui <- rxode2::rxode(readModelDb("Choi_2025_efpeglenatide"))

# Typical-value (no between-subject variability) copy, used for every
# deterministic replication below; the stochastic cohort is built later.
mod_t <- readModelDb("Choi_2025_efpeglenatide") |> rxode2::zeroRe()
```

* Citation: `r ui$reference`
* Description: `r ui$description`
* Article: <https://doi.org/10.3389/fphar.2025.1715585>

Efpeglenatide (HM11260C) is a long-acting GLP-1 receptor agonist: a
CA-exendin-4 analogue covalently linked to the Fc fragment of human IgG4
through a non-peptidyl linker. Choi 2025 is the first published population PK
analysis of the compound.

The distinguishing structural feature is a **dual subcutaneous absorption
pathway**. The observed concentration-time profiles show a double absorption
peak, which the authors described empirically by splitting each administered
dose into a fraction `BIOA1` deposited directly into the subcutaneous depot
(producing the first peak) and a complementary fraction
`BIOA2 = 1 - BIOA1` routed through a Savic 2007 transit-compartment chain
before discharging into the *same* depot (producing the delayed second peak).
Both fractions then absorb into the central compartment with the same
first-order rate constant `ka`.

# Population

Choi 2025 pooled 3,596 plasma efpeglenatide concentrations from 498
participants in six completed studies -- one phase 1 trial (HM-EXC-102) and
five phase 2 trials (HM-EXC-201, -202, -203, -204, -205); 3,316 post-dose
concentrations above the LLOQ entered the estimation. Five studies enrolled
adults with type 2 diabetes mellitus (293 participants, 58.8%) and one
(HM-EXC-205) enrolled non-diabetic adults with obesity (205 participants,
41.2%). Liraglutide comparator arms were excluded.

The pooled cohort had a median body weight of 93.6 kg (range 49.7-191 kg), a
median BMI of 33.4 kg/m^2 (range 19.2-57.7), a median age of 52 years, and was
55.2% female; 79.9% were Caucasian, 13.1% Black and 4.6% Asian (Choi 2025
Table 2). Subcutaneous dosing spanned single ascending doses of 2-100 ug/kg
and multiple-dose regimens of 0.3-6 mg once weekly (QW), 6-8 mg every two
weeks (Q2W) and 8-16 mg once monthly (QM) (Table 1).

The same information is available programmatically from the model's
`population` metadata:

```{r population}
str(ui$population)
```

# Source trace

Every `ini()` entry carries an in-file comment naming its source location in
`inst/modeldb/specificDrugs/Choi_2025_efpeglenatide.R`. They are collected
here for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` | ka = 0.006 1/h | Table 3, row `k a`, RSE 4.33% |
| `lcl` | CL/F = 0.032 L/h | Table 3, row `CL/F in obesity`, RSE 1.79% |
| `lvc` | Vc/F = 2.80 L | Table 3, row `V C /F`, RSE 7.39% |
| `lvp` | Vp/F = 3.96 L | Table 3, row `V P /F`, RSE 6.57% |
| `lq` | Q/F = 0.073 L/h | Table 3, row `Q/F`, RSE 11.14% |
| `logitffo` | BIOF = -1.11 | Table 3, row `BIOF`, RSE 27.59%; derived row `BIO A1` = 0.248 with footnote a |
| `lmtt` | MTT = 2.680 h | Table 3, row `MTT`, RSE 13.47% |
| `lnn` | N = 5.520 | Table 3, row `N`, RSE 22.28% |
| `e_wt_ka` | -0.927 | Table 3, row `Covariate effect of body weight on k a`, RSE 14.24% |
| `e_wt_cl` | 0.964 | Table 3, row `Covariate effect of body weight on CL/F`, RSE 6.75%; placement on CL/F confirmed by Results 3.1 and Discussion |
| `e_diab_cl` | 1.375 | Table 3: CL/F 0.044 L/h (T2DM) vs 0.032 L/h (obesity); the "38% higher clearance in T2DM" of the Discussion |
| `etalka` | 21.53% CV | Table 3 IIV column, row `k a`, RSE 28.48% |
| `etalcl` | 23.01% CV | Table 3 IIV column, row `CL/F in obesity`, RSE 5.60% |
| `etalvc` | 68.31% CV | Table 3 IIV column, row `V C /F`, RSE 18.28% |
| `etalvp` | 24.97% CV | Table 3 IIV column, row `V P /F`, RSE 27.11% |
| `etalmtt` | 46.27% CV | Table 3 IIV column, row `MTT`, RSE 37.14% |
| `etalogitffo` | 66.93% CV | Table 3 IIV column, row `BIO A1`, RSE 14.6% |
| `propSd` | 0.15 | Table 3, row `eps prop`, RSE 0.48% |
| Transit chain, `Ktr = (n+1)/MTT` | n/a | Equations 1-3 and the Figure 1 caption |
| `d/dt(depot)` dual input | n/a | Equation 4 (see Errata on the dropped `Ktr` factor) |
| `d/dt(central)`, `d/dt(peripheral1)` | n/a | Equations 5-6 |
| `BIOA1 = expit(BIOF)`, `BIOA2 = 1 - BIOA1` | n/a | Equation 7 |
| IIV exponential, `omega^2 = log(1 + CV^2)` | n/a | Equations 8-9 |
| Proportional residual error | n/a | Equation 10 |
| Power covariate model, median-normalized | n/a | Equation 11 |
| Categorical covariate indicator model | n/a | Equation 12 |

# Structural checks

These checks are deterministic -- they depend only on the transcribed
parameter values -- so they are asserted tightly. Each compares a quantity
derived from the packaged model against a number **printed in the paper**, so
a mis-transcribed clearance, volume, exponent or logit would turn one of them
red.

```{r structural}
th <- ui$theta

cl <- exp(th[["lcl"]]); vc <- exp(th[["lvc"]])
vp <- exp(th[["lvp"]]); q  <- exp(th[["lq"]])
ka <- exp(th[["lka"]])
mtt <- exp(th[["lmtt"]]); nn <- exp(th[["lnn"]])

# 1. BIOA1 from the logit parameter (Choi 2025 Eq. 7 and Table 3 footnote a).
bioa1 <- plogis(th[["logitffo"]])

# 2. Transit rate constant (Figure 1 caption / Eqs. 1-3).
ktr <- (nn + 1) / mtt

# 3. Steady-state volume (Discussion: "yielding a VSS ~ 6.8 L").
vss <- vc + vp

# 4. Terminal disposition half-life from the two-compartment macro-constants.
#    CL/F differs by disease stratum, so there are two of them: the obesity
#    reference (CL/F = 0.032 L/h) and T2DM (x 1.375 = 0.044 L/h).
thalf_d <- function(clv, vcv = vc, vpv = vp, qv = q) {
  k10 <- clv / vcv; k12 <- qv / vcv; k21 <- qv / vpv
  bsum <- k10 + k12 + k21
  log(2) / ((bsum - sqrt(bsum^2 - 4 * k10 * k21)) / 2) / 24
}
hl_obes <- thalf_d(cl)
hl_t2dm <- thalf_d(cl * th[["e_diab_cl"]])

structural <- tibble::tribble(
  ~Quantity,                            ~Model,     ~Paper, ~Source,
  "BIOA1 (bolus dose fraction)",        bioa1,      0.248,  "Table 3 derived row + footnote a",
  "BIOA2 (transit dose fraction)",      1 - bioa1,  0.752,  "Table 3 derived row + footnote b",
  "Ktr (1/h)",                          ktr,        2.433,  "(n+1)/MTT, Fig. 1 caption",
  "Vss = Vc + Vp (L)",                  vss,        6.8,    "Discussion, 'VSS ~ 6.8 L'",
  "Terminal half-life, obesity (days)", hl_obes,    6.5,    "Discussion, 'approximately 6-7 days'",
  "Terminal half-life, T2DM (days)",    hl_t2dm,    6.5,    "Discussion, 'approximately 6-7 days'"
)
structural$pct_diff <- 100 * (structural$Model - structural$Paper) / structural$Paper
knitr::kable(structural, digits = c(NA, 4, 4, NA, 2),
             caption = "Deterministic structural quantities vs the values printed by Choi 2025.")

# Number of terminal half-lives in the 4 weeks the paper says are needed to
# reach steady state on QW dosing ("steady state by ~4 weeks (~4-5 half-lives)").
n_hl_4wk <- 28 / c(hl_obes, hl_t2dm)

stopifnot(
  # BIOA1 / BIOA2 / Ktr are exact algebra on printed values -> very tight.
  abs(bioa1 - 0.248)       < 0.001,
  abs((1 - bioa1) - 0.752) < 0.001,
  abs(ktr - 2.433)         < 0.005,
  # Vss is quoted to two significant figures ("~ 6.8 L").
  abs(vss - 6.8) < 0.1,
  # The paper gives ONE half-life range for a model with two clearance
  # strata, so the check is that the two strata bracket that range and that
  # their midpoint lands on it. A mis-transcribed CL, Vc, Vp or Q moves these
  # by tens of percent and breaks the bracket.
  hl_t2dm < 6.5, hl_obes > 6.5,
  abs(mean(c(hl_obes, hl_t2dm)) - 6.5) < 0.75,
  # ... and the paper's own "~4-5 half-lives in 4 weeks" arithmetic.
  all(n_hl_4wk > 3.5), all(n_hl_4wk < 5.5)
)
```

The terminal half-life is `r round(hl_obes, 2)` days in the obesity stratum
and `r round(hl_t2dm, 2)` days in T2DM. Choi 2025 quotes a single
"approximately 6-7 days", which these two strata bracket (midpoint
`r round(mean(c(hl_obes, hl_t2dm)), 2)` days), and both sit within the
clinically reported 5.6-7.5 days that the paper cites from Yoon 2020. They
also reproduce the paper's own arithmetic that four weekly doses amount to
"approximately 4-5 half-lives"
(`r paste(sprintf("%.1f", rev(n_hl_4wk)), collapse = "-")` here).

Absorption (t1/2 = `r round(log(2) / ka / 24, 1)` days) is *faster* than
terminal disposition in both strata, so the terminal phase is
elimination-driven and no flip-flop interpretation applies.

## Absorption mass balance: the AUC identity

The dual-pathway encoding is the one place where an implementation slip would
silently change total exposure: the bolus arm is scaled by `f(depot)` while
the transit arm is scaled by the `bio` argument of `transit()`, and the two
must sum to exactly one dose. rxode2's `transit()` reads `podo()` as the
*unscaled* record amount, so the two arms are independent.

For any linear model the identity `AUC(0, inf) = Dose / CL` is **exact**, so
it is the sharpest available check on the whole absorption encoding: dropping
the `1 - ffo`, setting `f(depot) <- 1`, or omitting the leading `Ktr` from the
transit input rate (see the Errata) each break it by tens of percent. Note
that all of the vignette's other exposure checks are *ratios*, which cancel
exactly this class of error -- this is the one gate that catches it.

```{r mass-balance}
auc_identity <- function(diab) {
  ev <- rxode2::et(amt = 6, cmt = "depot") |>
    rxode2::et(seq(0, 20000, by = 4), cmt = "central")
  s <- rxode2::rxSolve(mod_t, ev, params = c(WT = 92, DIS_DIAB = diab),
                       atol = 1e-12, rtol = 1e-10, returnType = "data.frame")
  # Trapezoidal AUC over a window many half-lives long (Cc has decayed to
  # ~1e-9 of Cmax by 20000 h, so the unsampled tail is negligible).
  auc <- sum(diff(s$time) * (head(s$Cc, -1) + tail(s$Cc, -1)) / 2)
  clv <- cl * th[["e_diab_cl"]]^diab
  # Cc is ng/mL and dose is mg, so Dose/CL (mg / (L/h) = mg*h/L) x 1000.
  c(auc = auc, expected = 1000 * 6 / clv, ratio = auc / (1000 * 6 / clv))
}
ident <- rbind(obesity = auc_identity(0), T2DM = auc_identity(1))
knitr::kable(as.data.frame(ident), digits = c(0, 0, 6),
             caption = "AUC(0,inf) from the packaged model vs the exact Dose/CL identity.")

# Solver accuracy, not model accuracy, sets the achievable tolerance here
# (realised ~2e-5). 1e-3 keeps ample headroom while still catching any
# fractional-dose or missing-Ktr error, which are >10% effects.
stopifnot(max(abs(ident[, "ratio"] - 1)) < 1e-3)

# Independently, the total amount reaching the systemic circulation must equal
# the record amount: suppress elimination and distribution and read the plateau.
mod_mb <- readModelDb("Choi_2025_efpeglenatide") |>
  rxode2::zeroRe() |>
  rxode2::ini(lcl = log(1e-12), lq = log(1e-12))
ev_mb <- rxode2::et(amt = 10, cmt = "depot") |>
  rxode2::et(seq(0, 6000, by = 20), cmt = "central")
sim_mb <- rxode2::rxSolve(mod_mb, ev_mb, params = c(WT = 92, DIS_DIAB = 0),
                          returnType = "data.frame")
delivered <- max(sim_mb$central + sim_mb$peripheral1)
cat(sprintf("record amt = 10 mg; delivered to the systemic circulation = %.5f mg\n",
            delivered))
stopifnot(abs(delivered / 10 - 1) < 0.005)
```

## The double absorption peak

Choi 2025's motivation for the dual pathway was the double peak in the
absorption phase. The typical-value profile below reproduces it: an early
peak from the `BIOA1` bolus arm and a delayed shoulder from the transit arm.

```{r double-peak, fig.height = 3.6}
ev_pk <- rxode2::et(amt = 6, cmt = "depot") |>
  rxode2::et(seq(0, 400, by = 0.5), cmt = "central")
sim_pk <- rxode2::rxSolve(mod_t, ev_pk, params = c(WT = 92, DIS_DIAB = 0),
                          returnType = "data.frame")

ggplot(sim_pk, aes(time, Cc)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time (h)", y = "Cc (ng/mL)",
       title = "Typical-value profile, 6 mg SC single dose, 92 kg",
       caption = "Structure of Figure 1 of Choi 2025: bolus arm + delayed transit arm.")

# The transit arm (MTT 2.68 h, Ktr 2.43 /h) delivers its mass within the first
# few hours, so its imprint is a change of slope early in the absorption
# phase rather than a separate local maximum at these parameter values.
early <- dplyr::filter(sim_pk, time <= 40)
cat(sprintf("Concentration at 1 / 5 / 20 / 40 h: %.1f / %.1f / %.1f / %.1f ng/mL\n",
            early$Cc[early$time == 1], early$Cc[early$time == 5],
            early$Cc[early$time == 20], early$Cc[early$time == 40]))
stopifnot(all(diff(sim_pk$Cc[sim_pk$time <= 100]) > -1e-6))  # rises to peak
```

# Replicate published figures

## Figure 4A -- once-weekly dosing across 2-18 mg

```{r figure-4a}
qw_doses <- c(2, 4, 6, 8, 10, 12, 14, 16, 18)
tau <- 168          # QW, hours
n_wk <- 11

make_qw <- function(dose, wt, diab, id_offset = 0L, weeks = n_wk) {
  ev <- rxode2::et(amt = dose, cmt = "depot", time = seq(0, (weeks - 1) * tau, by = tau)) |>
    rxode2::et(seq(0, weeks * tau, by = 3), cmt = "central")
  d <- as.data.frame(ev)
  d$id <- id_offset + 1L
  d$WT <- wt
  d$DIS_DIAB <- diab
  d$dose <- dose
  d
}

ev_4a <- dplyr::bind_rows(
  lapply(seq_along(qw_doses), function(i)
    make_qw(qw_doses[i], wt = 92, diab = 1, id_offset = i - 1L))
)
stopifnot(!anyDuplicated(unique(ev_4a[, c("id", "time", "evid")])))

sim_4a <- rxode2::rxSolve(mod_t, ev_4a, keep = c("dose"),
                          returnType = "data.frame")

ggplot(sim_4a, aes(time / tau, Cc, colour = factor(dose))) +
  geom_line(linewidth = 0.6) +
  scale_colour_viridis_d(name = "Dose (mg QW)", option = "C", end = 0.9) +
  labs(x = "Time (weeks)", y = "Cc (ng/mL)",
       title = "Figure 4A -- once-weekly efpeglenatide, 2-18 mg",
       caption = "Replicates Figure 4A of Choi 2025.")
```

Choi 2025 reports "approximately dose-proportional increases" in
steady-state AUC across this range and that "steady-state concentrations were
generally achieved within 4 weeks at each dose level". Both are checked
below. Dose proportionality is exact for this model (elimination is linear
and `transit()` scales with the record amount), so the tolerance is tight and
would catch any dose-dependent leak in the absorption encoding.

```{r figure-4a-checks}
ss <- sim_4a |>
  dplyr::filter(time >= (n_wk - 1) * tau, time <= n_wk * tau) |>
  dplyr::group_by(dose) |>
  dplyr::summarise(cav_ss = mean(Cc), cmax_ss = max(Cc), .groups = "drop") |>
  dplyr::mutate(cav_per_mg = cav_ss / dose)

knitr::kable(
  ss |> dplyr::rename("Dose (mg QW)" = dose, "Cavg,ss (ng/mL)" = cav_ss,
                      "Cmax,ss (ng/mL)" = cmax_ss,
                      "Cavg,ss per mg" = cav_per_mg),
  digits = 2, caption = "Steady-state exposure over the final QW interval."
)

# Dose proportionality: dose-normalised Cavg,ss must be constant.
prop_spread <- diff(range(ss$cav_per_mg)) / mean(ss$cav_per_mg)
cat(sprintf("dose-normalised Cavg,ss spread across 2-18 mg: %.3g\n", prop_spread))
# Exactly zero in theory; the realised spread (~4e-7) is solver tolerance.
# 1e-4 keeps headroom over that while still catching any dose-dependent leak
# in the absorption encoding, which would be a percent-level effect.
stopifnot(prop_spread < 1e-4)

# Approach to steady state: trough at week 4 vs the final trough.
troughs <- sim_4a |>
  dplyr::filter(dose == 6, time %in% (seq_len(n_wk) * tau)) |>
  dplyr::arrange(time)
frac_wk4 <- troughs$Cc[troughs$time == 4 * tau] / troughs$Cc[nrow(troughs)]
cat(sprintf("week-4 trough is %.1f%% of the final trough\n", 100 * frac_wk4))
# Deterministic (zeroRe); with t1/2 = 7.1 d and tau = 7 d, 4 doses gives ~4
# half-lives. 0.85 leaves headroom below the realised value but still breaks
# if the clearance or volumes are mis-transcribed enough to move the
# accumulation profile materially.
stopifnot(frac_wk4 > 0.85, frac_wk4 <= 1)
```

## Figure 4B -- type 2 diabetes vs obesity

```{r figure-4b}
ev_4b <- dplyr::bind_rows(
  make_qw(6,  wt = 96, diab = 0, id_offset = 0L) |> dplyr::mutate(grp = "6 mg, obesity"),
  make_qw(6,  wt = 92, diab = 1, id_offset = 1L) |> dplyr::mutate(grp = "6 mg, T2DM"),
  make_qw(10, wt = 96, diab = 0, id_offset = 2L) |> dplyr::mutate(grp = "10 mg, obesity"),
  make_qw(10, wt = 92, diab = 1, id_offset = 3L) |> dplyr::mutate(grp = "10 mg, T2DM"),
  make_qw(16, wt = 96, diab = 0, id_offset = 4L) |> dplyr::mutate(grp = "16 mg, obesity"),
  make_qw(16, wt = 92, diab = 1, id_offset = 5L) |> dplyr::mutate(grp = "16 mg, T2DM")
)
stopifnot(!anyDuplicated(unique(ev_4b[, c("id", "time", "evid")])))

sim_4b <- rxode2::rxSolve(mod_t, ev_4b, keep = c("dose", "grp"),
                          returnType = "data.frame") |>
  dplyr::mutate(population = ifelse(grepl("obesity", grp), "Obesity", "T2DM"))

ggplot(sim_4b, aes(time / tau, Cc, colour = factor(dose), linetype = population)) +
  geom_line(linewidth = 0.6) +
  scale_colour_viridis_d(name = "Dose (mg QW)", option = "D", end = 0.85) +
  scale_linetype_manual(name = NULL, values = c(Obesity = "solid", T2DM = "dashed")) +
  labs(x = "Time (weeks)", y = "Cc (ng/mL)",
       title = "Figure 4B -- obesity vs T2DM, 6 / 10 / 16 mg QW",
       caption = "Replicates Figure 4B of Choi 2025.")
```

```{r figure-4b-ratio}
ratio_4b <- sim_4b |>
  dplyr::filter(time >= (n_wk - 1) * tau, time <= n_wk * tau) |>
  dplyr::group_by(dose, population) |>
  dplyr::summarise(cav = mean(Cc), .groups = "drop") |>
  tidyr::pivot_wider(names_from = population, values_from = cav) |>
  dplyr::mutate(ratio = Obesity / T2DM)

knitr::kable(
  ratio_4b |> dplyr::rename("Dose (mg QW)" = dose,
                            "Cavg,ss obesity (ng/mL)" = Obesity,
                            "Cavg,ss T2DM (ng/mL)" = T2DM,
                            "Obesity / T2DM" = ratio),
  digits = 3,
  caption = "Steady-state exposure ratio, obesity (96 kg) vs T2DM (92 kg)."
)

cat(sprintf("model obesity/T2DM exposure ratio: %.3f (paper Results 3.2 states ~1.21)\n",
            mean(ratio_4b$ratio)))
```

The model gives an obesity-to-T2DM steady-state exposure ratio of
`r sprintf("%.2f", mean(ratio_4b$ratio))` at the subgroup median weights
(1.375 exactly at a common body weight -- the "38% higher clearance in T2DM"
that Choi 2025's Discussion states). The paper's Results section instead
summarises Figure 4B as "approximately 21% higher", and Figure 4B's own
plotted curves read a ratio of about 1.24. This is an unreconciled
inconsistency **within the paper**, discussed in the Errata below; the
packaged model follows Table 3, which the Discussion's own effect-size
statement confirms exactly. It is therefore left out of the assertion gate
rather than accommodated by widening a tolerance.

## Figure 5 -- stepwise dose escalation

```{r figure-5}
esc_targets <- c(4, 6, 8, 10, 12)
esc_weeks <- 28

# 2 mg for 4 weeks, then +2 mg every 4 weeks until the target is reached,
# then the target dose for the remainder (Choi 2025 Figure 5 caption).
esc_amt <- function(target) {
  wk <- seq_len(esc_weeks) - 1L
  step <- 2 + 2 * (wk %/% 4)
  pmin(step, target)
}

ev_5 <- dplyr::bind_rows(c(
  lapply(seq_along(esc_targets), function(i) {
    tgt <- esc_targets[i]
    ev <- rxode2::et(amt = 1, cmt = "depot", time = (seq_len(esc_weeks) - 1L) * tau) |>
      rxode2::et(seq(0, esc_weeks * tau, by = 6), cmt = "central")
    d <- as.data.frame(ev)
    d$amt[d$evid == 1] <- esc_amt(tgt)
    d$id <- i
    d$WT <- 96; d$DIS_DIAB <- 0
    d$grp <- paste0("escalation to ", tgt, " mg")
    d
  }),
  list(
    {
      ev <- rxode2::et(amt = 2, cmt = "depot", time = (seq_len(esc_weeks) - 1L) * tau) |>
        rxode2::et(seq(0, esc_weeks * tau, by = 6), cmt = "central")
      d <- as.data.frame(ev); d$id <- length(esc_targets) + 1L
      d$WT <- 96; d$DIS_DIAB <- 0; d$grp <- "2 mg weekly"; d
    },
    {
      ev <- rxode2::et(amt = 2, cmt = "depot", time = 0) |>
        rxode2::et(seq(0, esc_weeks * tau, by = 6), cmt = "central")
      d <- as.data.frame(ev); d$id <- length(esc_targets) + 2L
      d$WT <- 96; d$DIS_DIAB <- 0; d$grp <- "2 mg single"; d
    }
  )
))
stopifnot(!anyDuplicated(unique(ev_5[, c("id", "time", "evid")])))

sim_5 <- rxode2::rxSolve(mod_t, ev_5, keep = c("grp"), returnType = "data.frame")

ggplot(sim_5, aes(time / tau, Cc, colour = grp)) +
  geom_line(linewidth = 0.6) +
  scale_colour_viridis_d(name = NULL, option = "C", end = 0.92) +
  labs(x = "Time (weeks)", y = "Cc (ng/mL)",
       title = "Figure 5 -- stepwise once-weekly dose escalation from 2 mg",
       caption = "Replicates Figure 5 of Choi 2025.")

# Exposure must increase monotonically with the escalation target (a
# deterministic, typical-value comparison, so exact ordering is safe here).
# Measured over the FINAL dosing interval: the 12 mg arm only reaches its
# maintenance dose at week 20, so at weeks 24-25 it is still ~1.4% short of
# steady state and dose-normalised exposure is not yet flat. By weeks 27-28
# every arm has plateaued (realised spread 0.18%; 0.36% at weeks 26-27).
esc_ss <- sim_5 |>
  dplyr::filter(grepl("escalation", grp),
                time >= (esc_weeks - 1) * tau, time <= esc_weeks * tau) |>
  dplyr::group_by(grp) |>
  dplyr::summarise(cav = mean(Cc), .groups = "drop") |>
  dplyr::mutate(target = as.numeric(sub(".*to ", "", sub(" mg", "", grp)))) |>
  dplyr::arrange(target)
stopifnot(all(diff(esc_ss$cav) > 0))

# At the plateau, exposure is proportional to the maintenance dose. 1% leaves
# headroom over the realised 0.18% while still catching a dose-dependent leak
# in the absorption encoding.
esc_spread <- diff(range(esc_ss$cav / esc_ss$target)) / mean(esc_ss$cav / esc_ss$target)
cat(sprintf("dose-normalised Cavg over the final interval: spread %.4f\n", esc_spread))
stopifnot(esc_spread < 0.01)
```

## Figure 6 -- covariate subgroup forest plot

Figure 6 of Choi 2025 reports simulated geometric mean ratios of steady-state
AUC and Cmax for twelve covariate subgroups of obese participants, relative
to the obese median-weight reference (96 kg). Crucially, the figure caption
states that **every** subgroup was simulated by fixing body weight at that
subgroup's representative median -- age, sex and race enter only through
weight, since none of them was retained in the final model. The subgroup
weights below are transcribed from the Figure 6 caption.

```{r figure-6-cohort}
fig6 <- tibble::tribble(
  ~subgroup,                              ~WT,
  "Weight = 79 kg (10th pct)",             79,
  "Weight = 119 kg (90th pct)",           119,
  "Male",                                 110,
  "Female",                                95,
  "Caucasian",                             98,
  "Black",                                106,
  "Asian",                                 83,
  "Native Hawaiian or Pacific Islander",  100,
  "Others",                               101,
  "Age <= 34 years",                      103,
  "Age 34-52 years",                       95,
  "Age >= 52 years",                       97
)
ref_wt <- 96   # obese median weight, Figure 6 caption

ev_6 <- dplyr::bind_rows(
  lapply(seq_len(nrow(fig6) + 1L), function(i) {
    wt <- if (i == 1L) ref_wt else fig6$WT[i - 1L]
    lab <- if (i == 1L) "Reference (96 kg)" else fig6$subgroup[i - 1L]
    d <- make_qw(6, wt = wt, diab = 0, id_offset = i - 1L, weeks = 21)
    d$subgroup <- lab
    d
  })
)
stopifnot(!anyDuplicated(unique(ev_6[, c("id", "time", "evid")])))

sim_6 <- rxode2::rxSolve(mod_t, ev_6, keep = c("subgroup", "WT"),
                         returnType = "data.frame")
```

Steady-state AUC0-tau and Cmax are computed with PKNCA over the final QW
dosing interval.

```{r figure-6-pknca}
sim_6_nca <- sim_6 |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, subgroup)

# Guarantee a time = 0 record per subject (extravascular pre-dose Cc = 0).
sim_6_nca <- dplyr::bind_rows(
  sim_6_nca,
  sim_6_nca |> dplyr::distinct(id, subgroup) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, subgroup, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

dose_6 <- ev_6 |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, subgroup)

conc_6 <- PKNCA::PKNCAconc(sim_6_nca, Cc ~ time | subgroup + id,
                           concu = "ng/mL", timeu = "h")
dose_6o <- PKNCA::PKNCAdose(dose_6, amt ~ time | subgroup + id, doseu = "mg")

ss_start <- max(dose_6$time)
int_6 <- data.frame(start = ss_start, end = ss_start + tau,
                    cmax = TRUE, auclast = TRUE, cav = TRUE)

res_6 <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_6, dose_6o, intervals = int_6))

nca_6 <- as.data.frame(res_6) |>
  dplyr::select(subgroup, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

ref_row <- dplyr::filter(nca_6, subgroup == "Reference (96 kg)")
gmr <- nca_6 |>
  dplyr::filter(subgroup != "Reference (96 kg)") |>
  dplyr::mutate(
    gmr_auc  = auclast / ref_row$auclast,
    gmr_cmax = cmax    / ref_row$cmax
  ) |>
  dplyr::left_join(fig6, by = "subgroup")
```

The published AUC ratios read off Figure 6 are compared below. The two
subgroups whose ratios Choi 2025 also prints numerically (79 kg and the
90th-percentile weight, with 90% CIs) are the strongest anchors; the
remaining ten are digitised from the figure and carry roughly +/- 0.01
reading uncertainty.

```{r figure-6-compare}
gmr <- gmr |>
  dplyr::mutate(
    published_auc = c(
      "Weight = 79 kg (10th pct)"           = 1.220,
      "Weight = 119 kg (90th pct)"          = 0.822,
      "Male"                                = 0.877,
      "Female"                              = 1.010,
      "Caucasian"                           = 0.985,
      "Black"                               = 0.918,
      "Asian"                               = 1.140,
      "Native Hawaiian or Pacific Islander" = 0.952,
      "Others"                              = 0.958,
      "Age <= 34 years"                     = 0.930,
      "Age 34-52 years"                     = 1.020,
      "Age >= 52 years"                     = 0.998
    )[subgroup],
    pct_diff = 100 * (gmr_auc - published_auc) / published_auc,
    # Closed form: AUC0-tau = Dose/CL, and CL scales as (WT/96)^0.964.
    closed_form = (WT / ref_wt)^-0.964
  ) |>
  dplyr::arrange(WT)

knitr::kable(
  gmr |>
    dplyr::select(subgroup, WT, gmr_auc, closed_form, published_auc, pct_diff, gmr_cmax) |>
    dplyr::rename("Subgroup" = subgroup, "Weight (kg)" = WT,
                  "AUC GMR (PKNCA)" = gmr_auc, "AUC GMR (closed form)" = closed_form,
                  "AUC GMR (Choi 2025 Fig. 6)" = published_auc,
                  "% diff" = pct_diff, "Cmax GMR (PKNCA)" = gmr_cmax),
  digits = c(NA, 0, 3, 3, 3, 1, 3),
  caption = "Figure 6 subgroup exposure ratios: simulated vs published."
)

cat(sprintf("median |%% diff| = %.2f%%; 90th pct = %.2f%%; max = %.2f%%\n",
            median(abs(gmr$pct_diff)), quantile(abs(gmr$pct_diff), 0.9),
            max(abs(gmr$pct_diff))))

stopifnot(
  # PKNCA AUC0-tau must agree with the closed form Dose/CL to solver accuracy.
  max(abs(gmr$gmr_auc / gmr$closed_form - 1)) < 0.01,
  # Against the figure: centre must be tight (a wrong reference weight or
  # exponent moves every row by tens of percent), and the envelope must
  # absorb figure-reading error on the ten digitised rows.
  abs(median(gmr$pct_diff)) < 2,
  max(abs(gmr$pct_diff)) < 5
)
```

All twelve subgroup ratios reproduce, with a median absolute difference of
`r sprintf("%.1f%%", median(abs(gmr$pct_diff)))`. Because every subgroup
differs from the reference only through body weight, this table is
effectively a twelve-point validation of the `e_wt_cl` exponent (0.964) and
the 92 kg reference weight simultaneously.

# PKNCA validation

A dense single-dose simulation gives the NCA parameters that can be compared
against the one absolute PK quantity Choi 2025 reports numerically -- the
half-life. Both disease strata are simulated, because CL/F (and therefore the
terminal half-life) differs between them.

```{r pknca-single}
sd_arms <- tibble::tribble(
  ~dose, ~diab, ~treatment,
      2,     1, "2 mg (T2DM)",
      6,     1, "6 mg (T2DM)",
     16,     1, "16 mg (T2DM)",
      6,     0, "6 mg (obesity)"
)

ev_sd <- dplyr::bind_rows(
  lapply(seq_len(nrow(sd_arms)), function(i) {
    ev <- rxode2::et(amt = sd_arms$dose[i], cmt = "depot") |>
      rxode2::et(c(seq(0, 48, by = 0.5), seq(49, 2400, by = 4)), cmt = "central")
    d <- as.data.frame(ev)
    d$id <- i; d$WT <- 92; d$DIS_DIAB <- sd_arms$diab[i]
    d$treatment <- sd_arms$treatment[i]
    d
  })
)
sim_sd <- rxode2::rxSolve(mod_t, ev_sd, keep = c("treatment"),
                          returnType = "data.frame")
# rxSolve omits `id` for a single-subject event table; restore it defensively
# so the PKNCA grouping below cannot silently collapse (failure pattern 8).
if (is.null(sim_sd$id)) sim_sd$id <- 1L
stopifnot(all(sim_sd$Cc >= 0))

sd_nca <- sim_sd |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)
sd_nca <- dplyr::bind_rows(
  sd_nca,
  sd_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

dose_sd <- ev_sd |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)

conc_sd <- PKNCA::PKNCAconc(sd_nca, Cc ~ time | treatment + id,
                            concu = "ng/mL", timeu = "h")
dose_sdo <- PKNCA::PKNCAdose(dose_sd, amt ~ time | treatment + id, doseu = "mg")

int_sd <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, auclast = TRUE,
  half.life = TRUE
)

res_sd <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_sd, dose_sdo, intervals = int_sd))
knitr::kable(as.data.frame(summary(res_sd)),
             caption = "Single-dose NCA of the typical 92 kg profiles, by disease stratum.")
```

## Comparison against published NCA

```{r nca-compare}
published <- tibble::tibble(
  treatment = sd_arms$treatment,
  # Choi 2025 Discussion: "The model-derived effective half-life of
  # efpeglenatide was approximately 6-7 days" (156 h at the 6.5-day
  # midpoint), "consistent with clinically reported values of 5.6-7.5 days"
  # (Yoon 2020). The paper gives one range for both strata and does not
  # break it down by dose, so the same reference applies to every arm.
  half.life = 6.5 * 24
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = res_sd,
  reference = published,
  by        = "treatment",
  units     = c(half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated vs published NCA. * differs from the reference by >20%.",
  align = c("l", "l", "r", "r", "r")
)
```

```{r nca-compare-check}
hl <- as.data.frame(res_sd) |>
  dplyr::filter(PPTESTCD == "half.life") |>
  dplyr::select(treatment, half_life_d = PPORRES) |>
  dplyr::mutate(half_life_d = half_life_d / 24,
                analytic_d = ifelse(grepl("obesity", treatment), hl_obes, hl_t2dm))
knitr::kable(
  hl |> dplyr::rename("Arm" = treatment, "NCA t1/2 (days)" = half_life_d,
                      "Analytic terminal t1/2 (days)" = analytic_d),
  digits = 3, caption = "PKNCA half-life vs the analytic terminal half-life."
)

stopifnot(
  # Dose-independent within a stratum (deterministic typical-value profiles).
  diff(range(hl$half_life_d[grepl("T2DM", hl$treatment)])) < 0.05,
  # Obesity must be the slower stratum: CL/F is 1.375x lower there. This is a
  # deterministic comparison, so the ordering is safe to assert.
  hl$half_life_d[hl$treatment == "6 mg (obesity)"] >
    max(hl$half_life_d[grepl("T2DM", hl$treatment)]),
  # PKNCA's automatic lambda.z window sits slightly inside the true terminal
  # phase here, because ka (0.006 /h) and beta are close enough that the two
  # exponential terms nearly cancel; the estimate runs ~5% long against the
  # analytic value. 15% absorbs that without hiding a transcription error.
  max(abs(hl$half_life_d / hl$analytic_d - 1)) < 0.15,
  # Every arm within the clinically reported 5.6-7.5 days that Choi 2025
  # cites from Yoon 2020, with a little room for the NCA window bias.
  all(hl$half_life_d > 5.4), all(hl$half_life_d < 8.0)
)
```

PKNCA's automatic terminal-window selection returns half-lives about 5%
longer than the analytic values, because `ka` (0.006 /h) sits close enough to
the disposition rate constant `beta` that the two exponential terms in the
oral solution largely cancel and the profile approaches log-linearity only
very slowly. The analytic values in the table above are the authoritative
ones, and both readings agree with the paper's stated window.

# Stochastic simulation and pcVPC-style check

Choi 2025's Figure 3 shows prediction-corrected VPCs by regimen. The cohort
below draws 200 participants per regimen -- weights sampled log-normally to
reproduce the Table 2 median of 93.6 kg and the 49.7-191 kg range -- and
plots the median with the 95% prediction interval, including residual error.

```{r vpc-cohort}
# set.seed() seeds R's RNG, not rxode2's simulation RNG (which is partitioned
# per solver thread). The cohort therefore differs across machines with
# different thread counts; every assertion below is written to hold for any
# cohort the model can produce.
set.seed(20251201)
n_arm <- 200

sample_wt <- function(n) {
  wt <- exp(rnorm(n, log(93.6), 0.30))
  pmin(pmax(wt, 49.7), 191)
}

make_vpc_arm <- function(label, dose, interval, n_dose, diab, tmax_h, id_offset) {
  subj <- tibble::tibble(
    id = id_offset + seq_len(n_arm),
    WT = sample_wt(n_arm),
    DIS_DIAB = diab,
    regimen = label
  )
  dosing <- subj |>
    tidyr::crossing(time = (seq_len(n_dose) - 1L) * interval) |>
    dplyr::mutate(amt = dose, evid = 1L, cmt = "depot")
  obs <- subj |>
    tidyr::crossing(time = seq(0, tmax_h, length.out = 60)) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
  dplyr::bind_rows(dosing, obs) |> dplyr::arrange(id, time, dplyr::desc(evid))
}

ev_vpc <- dplyr::bind_rows(
  make_vpc_arm("Single dose (T2DM)", 6, tau,  1, 1, 168,  id_offset =   0L),
  make_vpc_arm("QW",                 6, tau, 12, 1, 12 * tau, id_offset = 200L),
  make_vpc_arm("Q4W",               12, 4 * tau, 4, 1, 16 * tau, id_offset = 400L)
)
stopifnot(!anyDuplicated(unique(ev_vpc[, c("id", "time", "evid")])))

mod_iiv <- readModelDb("Choi_2025_efpeglenatide")
sim_vpc <- rxode2::rxSolve(mod_iiv, ev_vpc, keep = c("regimen", "WT"),
                           addDosing = FALSE, returnType = "data.frame")
```

```{r vpc-plot, fig.height = 3.8}
sim_vpc |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(regimen, time) |>
  dplyr::summarise(
    lo  = quantile(Cc, 0.025),
    mid = median(Cc),
    hi  = quantile(Cc, 0.975),
    .groups = "drop"
  ) |>
  ggplot(aes(time / 24, mid)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.25, fill = "firebrick") +
  geom_line(linewidth = 0.7, colour = "navy") +
  facet_wrap(~regimen, scales = "free") +
  labs(x = "Time (days)", y = "Cc (ng/mL)",
       title = "Median and 95% prediction interval by regimen",
       caption = "Structure of Figure 3 of Choi 2025 (simulated cohort; no observed data).")
```

The width of that band is what settles the reading of the residual-error
estimate. Choi 2025 Table 3 reports `eps prop = 0.15` without stating whether
0.15 is the proportional SD or the NONMEM `$SIGMA` variance. Read as a
variance the SD would be 0.387, whose residual contribution alone puts the
2.5th percentile at `1 - 1.96 * 0.387 = 0.24` of the prediction -- already far
below the lower edge of the paper's own pcVPC (Figure 3 panel C reads a
lower-edge-to-median ratio of about 0.53) *before* any IIV is added. The
packaged model therefore takes 0.15 as the SD, and the check below confirms
the resulting band is compatible with Figure 3 rather than far wider than it.

```{r vpc-band-check}
band <- sim_vpc |>
  dplyr::filter(!is.na(Cc), regimen == "QW", time > 6 * tau) |>
  dplyr::group_by(time) |>
  dplyr::summarise(lo = quantile(Cc, 0.025), mid = median(Cc), .groups = "drop") |>
  dplyr::mutate(ratio = lo / mid)

cat(sprintf("lower-95%%-PI / median over the QW plateau: %.3f (Figure 3 reads ~0.5)\n",
            median(band$ratio)))
# A variance reading of sigma would drive this ratio towards ~0.2; the SD
# reading keeps it near the ~0.5 the paper's pcVPC shows. The bound is wide
# enough to be cohort-independent but still separates the two readings, which
# differ by more than a factor of two.
stopifnot(median(band$ratio) > 0.33)
```

# Assumptions and deviations / Errata

**Structural encoding**

- **The transit input rate carries the leading `Ktr` that Equation 4 drops.**
  As printed, Equation 4's second term is
  `Dose * BIOA2 * (Ktr*t)^n * exp(-Ktr*t) / n!`, which has units of amount,
  not amount per time, and integrates to `Dose * BIOA2 / Ktr` instead of
  `Dose * BIOA2`. The paper's own Equations 1-2 give the correct form:
  Equation 1 makes the outflow from the last transit compartment `Ktr * a_n`
  and Equation 2 gives `a_n(t)`, so the depot input rate is
  `Ktr * a_n(t) = Dose * BIOA2 * Ktr * (Ktr*t)^n * exp(-Ktr*t) / n!` -- the
  Savic 2007 form, and exactly what `rxode2::transit()` computes. Encoding
  Equation 4 literally would scale the transit arm by `1/Ktr = 0.41`. The
  mass-balance check above gates this.
- **Exact gamma instead of Stirling's approximation.** Equation 3 evaluates
  `n!` for non-integer `n` with `sqrt(2*pi) * n^(n+0.5) * exp(-n)`. rxode2
  uses `lgamma(n+1)`. At `n = 5.52` Stirling gives `log(n!) = 5.6834` against
  the exact 5.6985, so the paper's kernel amplitude is 1.5% high and its
  transit arm delivers 0.763 rather than 0.752 of the dose. The exact form is
  used here because it conserves mass exactly; 1.5% is an order of magnitude
  below the residual error. Same choice as `Lee_2015_sumatriptan`.
- **One depot, one `ka`.** Both absorption arms discharge into the same depot
  and share `ka`, per the Discussion ("both pathways were described using the
  same first-order rate constant (ka = 0.006 h-1)"). This differs from the
  sibling `Lee_2015_sumatriptan` dual-pathway model, which uses two depots
  with two distinct rate constants.
- **Dose-record convention.** Supply **one** dose record targeting `depot`
  with the full administered amount. `f(depot) <- ffo` admits the bolus
  fraction and `transit(nn, mtt, 1 - ffo)` supplies the delayed fraction from
  the same (unscaled) record amount.
- **IIV on `BIOA1` is additive on the logit scale.** Equation 7's stated
  purpose is "to confine the estimated fraction values to the range between 0
  and 1 **and to allow implementation of interindividual variability**", so
  the eta sits on `BIOF`, not on `BIOA1` itself. Table 3's 66.93% CV is
  converted with the paper's own Equation 9 and applied on the logit scale.
- **Diagonal OMEGA, no IOV.** Results 3.1 reports "no statistically
  significant covariance identified between IIV terms" and that IOV "was
  evaluated but not found to be significant and was therefore not included in
  the final model". No IIV was reported on `Q/F` or `N`, so those etas are
  absent rather than set to a guessed value.

**Errata and internal inconsistencies in Choi 2025**

- **Table 3 definition cell for the weight-on-CL/F exponent.** The row
  "Covariate effect (theta) of body weight on CL/F" prints the definition
  `(V_C/F) x (WT/92)^theta`, copied from the `ka` row's template. The row
  label, Results 3.1 ("body weight influenced both ka and CL/F ... exponents
  of -0.927 and 0.964, respectively") and the Discussion ("a significant
  covariate on both ka and CL/F ... normalized to 92 kg; the exponents of
  -0.927 and 0.964") all place 0.964 on CL/F. Three statements against one
  cell; the model puts it on CL/F. The twelve-row Figure 6 reproduction above
  confirms this independently -- a weight exponent on Vc would leave
  steady-state AUC weight-independent and flatten every row to 1.0.
- **Results 3.2 swaps the two weight-subgroup GMR labels.** The text states
  that "the 10th percentile individual (79 kg) exhibited approximately 18%
  lower AUC and Cmax (GMR 0.82; 90% CI: 0.81-0.83), whereas the 90th
  percentile individual (120 kg) demonstrated about 20% higher exposure
  (GMR 1.22; 90% CI: 1.20-1.24)". Figure 6 shows the opposite assignment --
  79 kg at about 1.22 and the 90th-percentile weight at about 0.82 -- and the
  Discussion agrees with the figure ("For a change from 96 to 79 kg
  (~17.7% decrease), AUC increased by approximately 20.7% and CL/F decreased
  by approximately 17.1%"). A positive exponent on CL/F necessarily makes the
  lighter subject the *higher*-exposure one. Figure 6 and the Discussion are
  correct; the Results sentence has the two labels transposed. The model
  reproduces the Discussion's numbers exactly (20.67% AUC increase, 17.13%
  CL/F decrease for 96 -> 79 kg) and every Figure 6 row to within 1.1%.
- **Figure 6 weight label: 119 vs 120 kg.** The caption gives the
  90th-percentile weight as 119 kg while Results 3.2 says 120 kg. 119 kg is
  used here because it is the caption's simulation input and reproduces the
  printed 90% CI more closely.
- **Obesity-vs-T2DM exposure ratio.** Results 3.2 states that obese
  participants had "approximately 21% higher steady-state AUC compared to
  subjects with T2DM", but Table 3's two CL/F values imply 37.5% at a common
  body weight (the "38% higher clearance in T2DM" of the Discussion) and
  about 31% at the subgroup median weights. Figure 4B's own plotted curves
  read roughly 1.24. Closing the gap to 1.21 would require the T2DM
  simulation to have used a body weight near 85 kg, below every per-study
  T2DM median in Table 2 (88-97.3 kg). The packaged model follows Table 3,
  which the Discussion's effect-size statement confirms exactly; the 21%
  summary is not reproduced and is deliberately excluded from the assertion
  gate rather than accommodated by a widened tolerance.
- **Residual-error scale.** Table 3 reports `eps prop = 0.15` without stating
  whether this is the proportional SD or the `$SIGMA` variance. It is taken
  as the SD; the pcVPC band argument is given in the section above. Were it a
  variance, the proportional SD would be 38.7%.
- **`MTT` IIV precision.** Table 3 gives the RSE of the MTT IIV as 37.14%
  while Results 3.1 gives 35.88% for the same term. This affects only the
  reported precision, not the 46.27% CV point estimate that the model uses.
- **Table 2 internal arithmetic.** Two inconsistencies are reproduced as
  printed in the `population` metadata rather than silently corrected: the
  Total age range (18-64 years) is narrower than the HM-EXC-201 per-study
  range (47-75 years), and the race counts sum to 501 rather than 498 (the
  "Others" row prints 13 subjects at 1.92%, but 13/498 = 2.61%).
- **Specimen wording.** Methods 2.2 states that "Serum concentrations of
  efpeglenatide were quantified using a validated ELISA", while the same
  section, Section 2.7 and the axis labels of Figures 2-5 all say plasma.
  `compartmentData` records plasma as the dominant designation.

**Simulation assumptions in this vignette**

- Original observed concentrations are not publicly available, so no panel
  here overlays real data; the Figure 3 replication shows model-predicted
  intervals only.
- Figure 4A of Choi 2025 does not state which population it was simulated in.
  Its absolute concentrations are consistent with the T2DM clearance
  (0.044 L/h) rather than the obesity clearance, so `DIS_DIAB = 1` at 92 kg
  is used for that panel. The dose-proportionality and
  time-to-steady-state checks are unaffected by the choice.
- Figure 4B's per-population body weights are not given. The obese median
  (96 kg, from the Figure 6 caption) and the pooled T2DM median (92 kg, the
  model's reference weight) are used; the resulting ratio is reported rather
  than gated.
- Ten of the twelve Figure 6 AUC ratios were digitised from the plotted
  symbols and carry about +/- 0.01 reading uncertainty; the 79 kg and
  90th-percentile rows are printed numerically in Results 3.2 with 90% CIs.
  No parameter was adjusted to improve agreement.
- The screened-but-not-retained covariates (age, sex, race, BMI, lean body
  mass) are recorded in the model file's `covariatesDataExcluded` list. The
  paper's five-level race screen is mapped onto the four registered
  `RACE_*` canonicals; its "Native Hawaiian or Pacific Islander" level
  (2 participants) is folded into `RACE_OTHER` because no canonical exists
  for it and the whole race covariate was rejected by the stepwise search.
