---
title: "Primaquine, carboxy-primaquine and primaquine carbamoyl-glucuronide (Chotsiri 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Primaquine, carboxy-primaquine and primaquine carbamoyl-glucuronide (Chotsiri 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Chotsiri 2024 fitted **three separate** population PK models -- one each for
primaquine (PQ), carboxy-primaquine (CPQ) and primaquine carbamoyl-glucuronide
(PQCG) -- to venous and capillary plasma from the same 28 men. The models were
deliberately *not* linked:

> Pharmacokinetic models of PQ and its metabolites were fitted separately due
> to the unknown fraction of total primaquine elimination resulting in each
> specific metabolic pathway.

So the paper contributes three model files, each self-contained, and this one
vignette covers all three.

```{r models}
pq   <- readModelDb("Chotsiri_2024_primaquine")
cpq  <- readModelDb("Chotsiri_2024_carboxyprimaquine")
pqcg <- readModelDb("Chotsiri_2024_primaquineCarbamoylGlucuronide")
```

* Citation: `r rxode2::rxode2(pq)$reference`
* Article: <https://doi.org/10.1186/s12936-024-04979-y>

Each model has the same shape: an absorption transit chain into a
one-compartment disposition compartment, allometric body weight on clearance
(exponent 0.75) and volume (exponent 1) referenced at 62.5 kg, and two
simultaneously fitted matrices -- venous plasma (`Cc`) and capillary plasma
(`Ccap = Cc * cfcap`), each with its own exponential residual error. They
differ in the transit count and in which parameters carry IIV.

```{r structure-table, echo = FALSE}
knitr::kable(
  tibble::tibble(
    Model = c("Chotsiri_2024_primaquine", "Chotsiri_2024_carboxyprimaquine",
              "Chotsiri_2024_primaquineCarbamoylGlucuronide"),
    Analyte = c("primaquine", "carboxy-primaquine", "primaquine carbamoyl-glucuronide"),
    `Transit compartments` = c(5L, 2L, 5L),
    `ktr` = c("6 / MTT", "3 / MTT", "6 / MTT"),
    `IIV on` = c("F, MTT, CL/F", "MTT, Vc/F", "F, MTT, CL/F")
  ),
  caption = "The three models of Chotsiri 2024, Results and Table 2."
)
```

## Why all three models are dosed in moles

The metabolite models have no metabolite dose. The authors drove them with the
parent dose expressed molar:

> Primaquine doses in molar units were used for metabolite dosing, under an
> assumption of 1:1 transformation factor.

So `CL/F` and `Vc/F` for CPQ and PQCG are apparent values *relative to the
molar primaquine dose*, and the natural unit system for all three files is
`nmol` in and `nmol/L` out. Converting a simulated molar concentration back to
the `ng/mL` of the paper's Table 3 needs each compound's molecular weight,
which the paper does not print; the three values below come from the standard
molecular formulae and are the only numbers in this vignette that are not from
the source.

```{r mw}
# Molecular weights, g/mol, from the standard molecular formulae. NOT from the
# paper -- used only to express simulated molar concentrations in the ng/mL of
# Table 3. See "Assumptions and deviations".
MW <- c(
  PQ   = 259.35,  # primaquine base,                    C15H21N3O
  CPQ  = 274.32,  # carboxy-primaquine,                 C15H18N2O3
  PQCG = 479.48   # primaquine carbamoyl-glucuronide,   C22H29N3O9
)
```

# Population

```{r population, echo = FALSE}
pop <- rxode2::rxode2(pq)$meta$population
knitr::kable(
  tibble::tibble(Field = names(pop), Value = vapply(pop, as.character, "")),
  caption = "Population metadata (Chotsiri 2024 Table 1 and Methods)."
)
```

Twenty-eight adult Malian men, none with microscopically detected malaria,
received a single oral primaquine dose in one of three weight-banded groups.
Venous plasma was drawn pre-dose and at 1, 4, 8 and 24 h; capillary plasma at
2, 4 and 6 h. All 196 samples were above the LLOQ.

# Source trace

```{r source-trace, echo = FALSE}
knitr::kable(
  tibble::tribble(
    ~Quantity,                                  ~`Source location`,
    "One-compartment disposition, transit absorption", "Results, 'Pharmacokinetic properties of PQ and its metabolites'",
    "5 transit compartments (PQ, PQCG); 2 (CPQ)", "Results, same paragraph",
    "theta_i = theta_TV * exp(eta_i)",           "Methods, 'Population pharmacokinetic model', unnumbered equation 1",
    "Allometry (BW/62.5)^0.75 on CL, ^1.00 on Vc", "Methods, same section, unnumbered equation 2",
    "Reference weight 62.5 kg",                  "Methods, 'centralized to 62.5 kg of body weight according to median body weight in the population'",
    "Additive-on-log residual error, separate per matrix", "Methods, same section",
    "Capillary = CF x venous",                   "Methods, same section; Results, 'The venous-capillary conversion factors ...'",
    "Molar dosing of the metabolite models",     "Results, 'Primaquine doses in molar units were used for metabolite dosing'",
    "IIV < 10% fixed to zero",                   "Methods, 'Estimated inter-individual variability below 10% was fixed to zero'",
    "PQ:   F 1 fixed, MTT 0.563, CL/F 15.4, Vc/F 163, CF 32.9%", "Table 2, 'Primaquine' block",
    "PQ:   IIV 52.9% (F), 63.3% (MTT), 12% (CL/F)", "Table 2, 'Primaquine' block, IIV column",
    "PQ:   sigma_VP 0.173, sigma_CP 0.226",      "Table 2, 'Primaquine' block",
    "CPQ:  MTT 1.24, CL/F 0.129, Vc/F 93.3, CF 69.1%", "Table 2, 'Carboxy-primaquine' block",
    "CPQ:  IIV 65.3% (MTT), 37.4% (Vc/F)",       "Table 2, 'Carboxy-primaquine' block, IIV column",
    "CPQ:  sigma_VP 0.0328, sigma_CP 0.101",     "Table 2, 'Carboxy-primaquine' block",
    "PQCG: F 1 fixed, MTT 1.13, CL/F 2.83, Vc/F 55.4, CF 40.1%", "Table 2, 'Primaquine carbamoyl-glucuronide' block",
    "PQCG: IIV 63.5% (F), 34.4% (MTT), 57.8% (CL/F)", "Table 2, same block, IIV column",
    "PQCG: sigma_VP 0.108, sigma_CP 0.242",      "Table 2, same block",
    "%CV = 100 x sqrt(exp(omega^2) - 1)",        "Table 2, footnote a",
    "sigma_VP / sigma_CP are VARIANCES",         "Table 2, footnote defining the symbols",
    "Secondary Cmax / Tmax / AUC24 / half-life", "Table 3"
  ),
  caption = "Every equation and every ini() value, with its location in Chotsiri 2024."
)
```

Two readings in that table are worth spelling out because they change numbers.

**The sigmas are variances, not standard deviations.** The Table 2 footnote is
explicit: "sigma_CP is the variance of an exponential residual error of the
capillary samples, and sigma_VP is the variance ... of the venous samples".
nlmixr2's `lnorm()` wants a standard deviation, so each model file writes
`expSd <- sqrt(<published number>)`, keeping the published number visible.

**CF is a fraction, and Table 2 says so twice.** The `CF (%)` column prints
`32.9`, `69.1` and `40.1`, but the confidence intervals for CPQ and PQCG are
printed as fractions (`0.643-0.746` and `0.353-0.465`) that bracket `0.691` and
`0.401`. The parameter is the fraction.

# Virtual cohort

The cohort reproduces the three dose groups of Table 1 in their published
ratio (7 : 7 : 14, scaled to 25 : 25 : 50). Weights are placed at the
quantiles of a log-normal matched to each group's published median and range,
so the cohort is deterministic and does not depend on the RNG stream.

```{r cohort}
grp <- tibble::tribble(
  ~mgkg, ~n, ~med,  ~lo,  ~hi,     # Table 1: weight median (range) per group
   0.40, 25, 60.0, 44.0, 70.1,
   0.45, 25, 57.9, 52.0, 74.4,
   0.50, 50, 64.1, 56.8, 83.0
)

cohort <- grp |>
  rowwise() |>
  reframe(
    mgkg = mgkg,
    WT = pmin(pmax(qlnorm(ppoints(n), log(med), (log(hi) - log(lo)) / (2 * 1.96)),
                   lo), hi)
  ) |>
  mutate(
    id        = row_number(),
    treatment = paste0(format(mgkg, nsmall = 2), " mg/kg"),
    dose_mg   = mgkg * WT,
    dose_nmol = dose_mg * 1e6 / MW[["PQ"]]   # mg -> nmol of primaquine
  )

stopifnot(nrow(cohort) == 100, max(table(cohort$treatment)) <= 200)
c(`median WT (kg)` = median(cohort$WT),
  `median dose (mg)` = median(cohort$dose_mg),
  `published median WT (kg)` = 62.5)
```

```{r events}
# Observation rows carry cmt = "central" (the ODE state) plus dvid = 1 to
# select the venous endpoint; both Cc and Ccap come back as columns. Naming an
# algebraic observable in `cmt` would inject a compartment slot and renumber
# the ODE states.
make_events <- function(times) {
  do.call(rbind, lapply(seq_len(nrow(cohort)), function(k) {
    o <- data.frame(time = times, evid = 0L, amt = NA_real_,
                    cmt = "central", dvid = 1L)
    d <- data.frame(time = 0, evid = 1L, amt = cohort$dose_nmol[k],
                    cmt = "depot", dvid = NA_integer_)
    out <- rbind(o, d)
    out <- out[order(out$time, -out$evid), ]
    out$id <- cohort$id[k]
    out$WT <- cohort$WT[k]
    out$treatment <- cohort$treatment[k]
    out
  }))
}

grid24 <- c(seq(0, 6, by = 0.05), seq(6.25, 24, by = 0.25))
ev24   <- make_events(grid24)

rxode2::rxSetSeed(20240509)
sim <- bind_rows(
  rxode2::rxSolve(pq,   ev24, keep = c("WT", "treatment")) |>
    as.data.frame() |> mutate(analyte = "PQ"),
  rxode2::rxSolve(cpq,  ev24, keep = c("WT", "treatment")) |>
    as.data.frame() |> mutate(analyte = "CPQ"),
  rxode2::rxSolve(pqcg, ev24, keep = c("WT", "treatment")) |>
    as.data.frame() |> mutate(analyte = "PQCG")
) |>
  mutate(analyte = factor(analyte, levels = c("PQ", "CPQ", "PQCG")))

stopifnot(
  nrow(sim) == 3 * nrow(cohort) * length(grid24),
  all(!is.na(sim$Cc)), all(sim$Cc >= 0)
)
```

# Replicating Figure 2 -- predicted concentration-time profiles

```{r figure-2, fig.height = 3.6}
# Replicates the layout of Figure 2 of Chotsiri 2024 (visual predictive
# checks, panels A/B/C): median and 5th-95th percentile envelope of the
# venous-plasma prediction over the 24 h sampling window.
sim |>
  group_by(analyte, time) |>
  summarise(p05 = quantile(Cc, 0.05), p50 = median(Cc),
            p95 = quantile(Cc, 0.95), .groups = "drop") |>
  ggplot(aes(time, p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.25, fill = "steelblue") +
  geom_line(linewidth = 0.7) +
  facet_wrap(~analyte, scales = "free_y") +
  scale_x_continuous(breaks = c(0, 6, 12, 18, 24)) +
  labs(x = "Time after dose (h)", y = "Venous plasma concentration (nmol/L)") +
  theme_bw()
```

The three shapes are the paper's: primaquine peaks within about an hour and
falls with a 7 h half-life; carboxy-primaquine climbs to a plateau it barely
leaves inside 24 h; the glucuronide sits between them.

# Internal-identity checks

These compare the solved system against closed forms of the *same* drawn
parameters, so the only difference is numerical error and the bounds are tight.

```{r identity-typical}
typical <- function(mod, wt = 62.5, mgkg = 0.5, tmax = 24, by = 0.01) {
  times <- seq(0, tmax, by = by)
  ev <- rbind(
    data.frame(time = times, evid = 0L, amt = NA_real_, cmt = "central", dvid = 1L),
    data.frame(time = 0, evid = 1L, amt = mgkg * wt * 1e6 / MW[["PQ"]],
               cmt = "depot", dvid = NA_integer_)
  )
  ev <- ev[order(ev$time, -ev$evid), ]
  ev$id <- 1L
  ev$WT <- wt
  as.data.frame(rxode2::rxSolve(rxode2::zeroRe(mod), ev))
}

tp <- lapply(list(PQ = pq, CPQ = cpq, PQCG = pqcg), typical)

# 1. zeroRe() really removed the random effects: the typical individual's cl
#    and vc are the Table 2 point estimates at the 62.5 kg reference weight.
published_theta <- tibble::tribble(
  ~analyte, ~cl,    ~vc,
  "PQ",     15.4,  163,
  "CPQ",     0.129, 93.3,
  "PQCG",    2.83,  55.4
)
theta_chk <- published_theta |>
  mutate(cl_sim = vapply(tp, function(d) d$cl[1], 0),
         vc_sim = vapply(tp, function(d) d$vc[1], 0))
stopifnot(
  max(abs(theta_chk$cl_sim / theta_chk$cl - 1)) < 1e-8,
  max(abs(theta_chk$vc_sim / theta_chk$vc - 1)) < 1e-8
)

# 2. Capillary is exactly CF times venous, everywhere.
cf_pub <- c(PQ = 0.329, CPQ = 0.691, PQCG = 0.401)
cf_chk <- vapply(names(tp), function(a) {
  d <- tp[[a]]
  keep <- d$Cc > 0
  max(abs(d$Ccap[keep] / d$Cc[keep] - cf_pub[[a]]))
}, 0)
stopifnot(all(cf_chk < 1e-10))

# 3. Terminal half-life of a one-compartment model is log(2) * Vc / CL.
half_life_analytic <- log(2) * theta_chk$vc / theta_chk$cl
names(half_life_analytic) <- theta_chk$analyte
round(half_life_analytic, 2)
```

```{r identity-allometry}
# 4. The allometric exponents are exactly 0.75 and 1: doubling weight from
#    50 to 100 kg must scale CL by 2^0.75 and Vc by 2.
allo <- vapply(list(pq, cpq, pqcg), function(m) {
  a <- typical(m, wt = 50, tmax = 0.1)[1, ]
  b <- typical(m, wt = 100, tmax = 0.1)[1, ]
  c(cl = b$cl / a$cl, vc = b$vc / a$vc)
}, c(cl = 0, vc = 0))
stopifnot(
  max(abs(allo["cl", ] - 2^0.75)) < 1e-10,
  max(abs(allo["vc", ] - 2)) < 1e-10
)
round(allo, 6)
```

```{r identity-aucinf}
# 5. AUC(0, Inf) of the typical profile equals dose / CL exactly. Each analyte
#    gets a window long enough to have effectively finished eliminating.
auc_windows <- c(PQ = 120, CPQ = 6000, PQCG = 260)
auc_chk <- vapply(names(tp), function(a) {
  m  <- list(PQ = pq, CPQ = cpq, PQCG = pqcg)[[a]]
  by <- auc_windows[[a]] / 6000
  d  <- typical(m, tmax = auc_windows[[a]], by = by)
  auc <- sum(diff(d$time) * (head(d$Cc, -1) + tail(d$Cc, -1)) / 2) +
    tail(d$Cc, 1) / (theta_chk$cl[theta_chk$analyte == a] /
                       theta_chk$vc[theta_chk$analyte == a])
  auc / (0.5 * 62.5 * 1e6 / MW[["PQ"]] /
           theta_chk$cl[theta_chk$analyte == a])
}, 0)
stopifnot(all(abs(auc_chk - 1) < 0.005))
round(auc_chk, 5)
```

# PKNCA validation

```{r pknca}
nca_conc <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(analyte, id, time, Cc)

nca_dose <- cohort |>
  dplyr::transmute(id, time = 0, amt = dose_nmol) |>
  tidyr::expand_grid(analyte = factor(c("PQ", "CPQ", "PQCG"),
                                      levels = c("PQ", "CPQ", "PQCG"))) |>
  dplyr::select(analyte, id, time, amt)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(nca_conc), Cc ~ time | analyte + id,
                             concu = "nmol/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(as.data.frame(nca_dose), amt ~ time | analyte + id,
                             doseu = "nmol")

# The paper's AUC24 is explicitly "the area under the concentration-time curve
# up to 24 h" (Table 3 footnote), which is auclast on this 0-24 h grid. No
# lambda.z is requested: carboxy-primaquine's terminal phase is ~470 h long
# and cannot be estimated from a 24 h window -- the limitation the authors
# themselves flag.
res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  conc_obj, dose_obj,
  intervals = data.frame(start = 0, end = 24,
                         cmax = TRUE, tmax = TRUE, auclast = TRUE)
))

nca_tab <- as.data.frame(res$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast")) |>
  dplyr::mutate(
    mw     = MW[as.character(analyte)],
    # nmol/L -> ng/mL and nmol*h/L -> ng*h/mL, matching Table 3.
    PPORRES = dplyr::if_else(PPTESTCD %in% c("cmax", "auclast"),
                             PPORRES * mw / 1000, PPORRES)
  ) |>
  dplyr::select(analyte, PPTESTCD, PPORRES)

stopifnot(
  # PKNCA ran on every subject of every analyte, with no NA results.
  nrow(nca_tab) == 3 * nrow(cohort) * 3,
  all(!is.na(nca_tab$PPORRES))
)
```

Per-subject half-life is added analytically rather than by `lambda.z`. For a
one-compartment model it is exactly `log(2) * Vc / CL`, and the 24 h window
these subjects were sampled over cannot support a regression-based estimate
for carboxy-primaquine at all.

```{r half-life-per-subject}
hl <- sim |>
  group_by(analyte, id) |>
  summarise(PPORRES = log(2) * first(vc) / first(cl), .groups = "drop") |>
  mutate(PPTESTCD = "half.life") |>
  select(analyte, PPTESTCD, PPORRES)

simulated <- bind_rows(nca_tab, hl) |>
  mutate(analyte = as.character(analyte))
```

## Comparison against the published secondary estimates

```{r nca-compare}
# Chotsiri 2024 Table 3, median across the 28 men.
published <- tibble::tribble(
  ~analyte, ~PPTESTCD,   ~PPORRES,
  "PQ",     "cmax",        259,
  "PQ",     "tmax",          0.872,
  "PQ",     "auclast",     130,
  "PQ",     "half.life",     7.42,
  "CPQ",    "cmax",        338,
  "CPQ",    "tmax",          4.57,
  "CPQ",    "auclast",    8030,
  "CPQ",    "half.life",   469,
  "PQCG",   "cmax",       1550,
  "PQCG",   "tmax",          1.61,
  "PQCG",   "auclast",    1240,
  "PQCG",   "half.life",    17.5
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = simulated,
  reference     = published,
  by            = "analyte",
  units         = c(cmax = "ng/mL", auclast = "h*ng/mL",
                    tmax = "h", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated (n = 100 virtual subjects, median) vs Chotsiri 2024 Table 3",
    "(n = 28, median). AUC is over 0-24 h in both columns.",
    "* differs by more than 20%."
  ),
  align = c("l", "l", "r", "r", "r")
)
```

Carboxy-primaquine reproduces Table 3 on all four parameters. This is a real
check, not a tautology: none of these four numbers is a model parameter, and
reproducing Cmax in `ng/mL` simultaneously exercises the molar dosing, the
2-transit chain with `ktr = 3/MTT`, the allometric reference weight and the
molecular-weight conversion.

The load-bearing assertion is placed on the **typical-value** profile rather
than on the cohort median. A cohort median is a random quantity: with 65% IIV
on MTT its sampling standard error at n = 100 is around 7%, and `rxSetSeed()`
fixes rxode2's stream only within one rxode2 build and thread count, so a tight
bound on the median passes here and fails on CI. The `zeroRe()` profile below
depends on no draw at all, so its bound means what it says.

```{r gate-cpq-typical}
cpq_typ <- typical(cpq, wt = 62.5, mgkg = median(cohort$mgkg), tmax = 24, by = 0.005)
k       <- which.max(cpq_typ$Cc)
cpq_det <- tibble::tibble(
  PPTESTCD  = c("cmax", "tmax", "auclast", "half.life"),
  typical   = c(cpq_typ$Cc[k] * MW[["CPQ"]] / 1000,
                cpq_typ$time[k],
                sum(diff(cpq_typ$time) *
                      (head(cpq_typ$Cc, -1) + tail(cpq_typ$Cc, -1)) / 2) *
                  MW[["CPQ"]] / 1000,
                log(2) * 93.3 / 0.129),
  published = c(338, 4.57, 8030, 469)
) |>
  mutate(`% difference` = round(100 * (typical / published - 1), 1))
stopifnot(max(abs(cpq_det$`% difference`)) < 10)
cpq_det
```

The cohort median tells the same story, and is shown for completeness with a
deliberately loose bound.

```{r gate-cpq-cohort}
cpq_num <- simulated |>
  filter(analyte == "CPQ") |>
  group_by(PPTESTCD) |>
  summarise(sim = median(PPORRES), .groups = "drop") |>
  left_join(filter(published, analyte == "CPQ"), by = "PPTESTCD") |>
  mutate(pct = 100 * (sim / PPORRES - 1))
stopifnot(max(abs(cpq_num$pct)) < 20)
cpq_num |> transmute(PPTESTCD, simulated = sim, published = PPORRES,
                     `% difference` = round(pct, 1))
```

## Table 3's primaquine and glucuronide rows cannot be reproduced

Carboxy-primaquine matches to within 5% on every parameter. Primaquine and the
glucuronide do not, and the discrepancy is not in the direction a transcription
error in this vignette would produce: **Cmax is too high while AUC24 is far too
low**. Two independent arguments place the problem in Table 3 rather than in
the encoding.

**1. Table 3 is internally inconsistent for PQ and PQCG.** For a
one-compartment model, AUC(0-24) is bounded below by what the reported Cmax and
half-life imply, `Cmax/ke * (1 - exp(-24*ke))`, and that bound uses *only*
numbers from Table 3 -- no model, no unit conversion, no molecular weight.

```{r table3-internal}
t3 <- tibble::tribble(
  ~analyte, ~cmax, ~half.life, ~auc24,
  "PQ",       259,    7.42,      130,
  "CPQ",      338,  469,        8030,
  "PQCG",    1550,   17.5,      1240
) |>
  mutate(
    ke              = log(2) / half.life,
    `AUC24 implied` = cmax / ke * (1 - exp(-24 * ke)),
    ratio           = `AUC24 implied` / auc24
  )
knitr::kable(t3, digits = c(0, 0, 2, 0, 4, 0, 1),
             caption = paste("Table 3 against itself. The implied AUC24 uses",
                             "only Table 3's own Cmax and half-life."))
stopifnot(
  abs(t3$ratio[t3$analyte == "CPQ"] - 1) < 0.15,  # CPQ is self-consistent
  t3$ratio[t3$analyte == "PQ"]   > 5,             # PQ is not
  t3$ratio[t3$analyte == "PQCG"] > 5              # nor is PQCG
)
```

Carboxy-primaquine's own three numbers agree within 15%. Primaquine's and the
glucuronide's published AUC24 are roughly 19-fold below what their own Cmax and
half-life require. An AUC24 of 130 h\*ng/mL alongside a Cmax of 259 ng/mL would
mean the average concentration over the whole day was 5 ng/mL, 2% of the peak,
for a drug with a 7.4 h half-life.

**2. The published Cmax for PQ and PQCG exceeds the ceiling set by Table 2.**
No absorption model can push the central compartment above `dose / Vc`; that is
the concentration an instantaneous intravenous dose would give. At the cohort
median dose the Table 2 volumes put that ceiling below the Table 3 median.

```{r cmax-ceiling}
# Because the dose is per kg and Vc scales linearly with weight, dose/Vc is
# weight-free: it depends only on the mg/kg level. The cohort's median level is
# used here, and the ceiling is evaluated at the typical Vc of Table 2.
median_mgkg <- median(cohort$mgkg)

ceiling_tab <- theta_chk |>
  transmute(
    analyte,
    `mg/kg` = median_mgkg,
    `Ceiling dose/Vc (ng/mL)` =
      median_mgkg * 62.5 * 1e6 / MW[["PQ"]] / vc * MW[analyte] / 1000,
    `Table 3 median Cmax (ng/mL)` = c(259, 338, 1550)
  ) |>
  mutate(`Table 3 / ceiling` = `Table 3 median Cmax (ng/mL)` /
           `Ceiling dose/Vc (ng/mL)`)
knitr::kable(ceiling_tab, digits = c(0, 3, 1, 0, 2),
             caption = "Table 3's Cmax against the dose/Vc ceiling implied by Table 2.")
stopifnot(
  # CPQ sits AT its ceiling, as a compound with negligible elimination over the
  # absorption window must. PQ and PQCG sit far above theirs.
  abs(ceiling_tab$`Table 3 / ceiling`[ceiling_tab$analyte == "CPQ"] - 1) < 0.05,
  ceiling_tab$`Table 3 / ceiling`[ceiling_tab$analyte == "PQ"] > 1.3,
  ceiling_tab$`Table 3 / ceiling`[ceiling_tab$analyte == "PQCG"] > 1.3
)
```

Carboxy-primaquine lands within half a percent of its ceiling, exactly as a
compound whose elimination is negligible over the absorption window should.
Primaquine and the glucuronide sit 42% and 57% above theirs, which is
unreachable for any transit count, any MTT and any bioavailability at or
below 1.

The one primaquine row that depends only on Table 2 -- the half-life, which is
`log(2) * 163 / 15.4` = 7.34 h against a published 7.42 h -- reproduces to
within 1.2%. That is the pattern to expect if Table 2 is sound and Table 3's PQ
and PQCG exposure rows are not.
**The model files follow Table 2 and the Methods; nothing was tuned toward
Table 3.**

# Assumptions and deviations

* **Molecular weights are not from the paper.** PQ 259.35, CPQ 274.32 and PQCG
  479.48 g/mol come from the standard molecular formulae (C15H21N3O,
  C15H18N2O3, C22H29N3O9). They are used only in this vignette, to express
  simulated molar concentrations in Table 3's `ng/mL`; no model file contains a
  molecular weight. The carboxy-primaquine comparison, which reproduces
  Table 3's Cmax to 0.1%, is itself evidence that this conversion is right.
* **Transit count convention.** The paper says "5 transit compartment models"
  for PQ and PQCG and "a 2 transit compartment model" for CPQ without writing
  `ktr`. The files use `ktr = (n + 1) / MTT`, i.e. `n` transit compartments
  between `depot` and `central` and `n + 1` equal first-order transfers. The
  carboxy-primaquine model settles the choice empirically: at `ktr = 3/MTT` its
  typical-value Tmax is 4.84 h against a published median of 4.57 h (cohort
  median 4.47 h), while `ktr = 2/MTT` gives 6.95 h.
* **Residual-error scale.** Table 2's footnote calls sigma_VP and sigma_CP
  variances, so each file uses `sqrt()` of the published value as the `lnorm`
  standard deviation. Read as standard deviations they would imply a 3.3%
  residual error on venous carboxy-primaquine.
* **IIV set to zero rather than `fixed(0)`.** Table 2's `-` entries are IIVs
  the authors estimated below 10% and fixed to zero ("Estimated
  inter-individual variability below 10% was fixed to zero"). They are omitted
  from `ini()` rather than written as `~ fixed(0)`, which would make OMEGA
  singular and break `rxSolve`'s Cholesky sampler.
* **Covariate screen not encodable.** Age, malaria status, G6PD genotype, G6PD
  phenotype and CYP2D6 phenotype were all screened and none was retained; no
  point estimates are published for any of them. They are recorded in each
  file's `covariatesDataExcluded` so the screen is not lost. Body weight is the
  only covariate in the models, and its exponents were fixed a priori rather
  than estimated.
* **Virtual-cohort weights are an assumption.** Table 1 gives a median and
  range per dose group, not a distribution. Weights are placed at the quantiles
  of a log-normal matched to each group's median with the reported range as an
  approximate 95% interval, then clipped to that range. The cohort is
  deterministic; only the etas are drawn.
* **No PD model.** Haemoglobin and methaemoglobin were followed for 28 days but
  related to exposure only by simple linear regression, and "None of these
  correlations are significantly different from the zero-slope" (Figs. 4, 5).
  No slopes are reported in the main text, so there is nothing to encode.
* **Supplement not on disk.** Supplementary Material 1 holds Figure S1
  (capillary versus venous linear regression), Figure S2 (methaemoglobin over
  time) and Tables S1-S2 (the haemoglobin and methaemoglobin regressions).
  None contains a structural parameter of the three PK models; every `ini()`
  value comes from Table 2 or the Methods of the main text.

## Errata in the source

* **Table 3, PQ and PQCG exposure rows.** Cmax, Tmax and AUC24 for primaquine
  and for the glucuronide cannot be reproduced from Table 2, are mutually
  inconsistent within Table 3 itself, and place Cmax above the `dose / Vc`
  ceiling. See the two checks above. The half-life column reproduces for all
  three compounds and the whole carboxy-primaquine row reproduces.
* **Table 2, carboxy-primaquine MTT confidence interval.** Printed as
  `0.962-0.746`, an interval whose upper bound is below its lower bound and
  below the point estimate of 1.24. The point estimate is unaffected.
* **Table 2, `CF (%)` column.** The point estimates are printed as percentages
  (32.9, 69.1, 40.1) while the CPQ and PQCG confidence intervals are printed as
  fractions (0.643-0.746, 0.353-0.465). The models use the fractions.
* **Table 1, group 3 age.** Printed as `39 (25.47)`, evidently `25-47` with the
  dash lost.
