---
title: "PFOS gestational and lactational PBPK in rats and humans (Chou and Lin 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PFOS gestational and lactational PBPK in rats and humans (Chou and Lin 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

```{r libraries, message = FALSE}
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Chou and Lin (2021, *Environ Health Perspect* 129:037004,
[doi:10.1289/EHP7671](https://doi.org/10.1289/EHP7671)) developed a
life-stage physiologically based pharmacokinetic (PBPK) model for
**perfluorooctane sulfonate (PFOS)** spanning the pre-pregnant, pregnant
and lactating female, in both the Sprague-Dawley rat and the human. The
distinctive feature relative to earlier PFOS models is an explicit,
physiologically based description of transporter-mediated renal
reabsorption (saturable basolateral Oat1/Oat3 and apical Oatp1a1 uptake
in the proximal tubule cells) carried through gestation and lactation.
The rat model was calibrated against toxicokinetic studies (Thibodeaux
et al. 2003; Chang et al. 2009; Luebker et al. 2005a, 2005b) and the
human model against worldwide maternal-plasma, cord-blood, placental and
breast-milk biomonitoring data.

The paper contributes **six** models, extracted here as six files, all
sharing the same maternal core structure:

| Model file | Role |
| --- | --- |
| `Chou_2021_pfos_rat_prepregnant_pbpk` | Non-pregnant female rat; seeds the gestational run |
| `Chou_2021_pfos_rat_gestational_pbpk` | Pregnant rat + litter (placenta, fetus, amniotic fluid) |
| `Chou_2021_pfos_rat_lactational_pbpk` | Lactating dam + nursing pups (milk transfer, coprophagy) |
| `Chou_2021_pfos_human_prepregnant_pbpk` | Non-pregnant woman; 30-year pre-conception burn-in |
| `Chou_2021_pfos_human_gestational_pbpk` | Pregnant woman + fetus |
| `Chou_2021_pfos_human_lactational_pbpk` | Lactating woman + breastfed neonate |

The maternal core is a two-compartment gastrointestinal tract (stomach,
small intestine) feeding the liver by the portal vein; flow-limited
plasma, liver, fat, mammary gland and rest-of-body compartments in which
only the plasma-unbound fraction exchanges; and a three-subcompartment
kidney (kidney blood, proximal tubule cells, filtrate) combining
glomerular filtration with saturable basolateral and apical
reabsorption, passive diffusion and first-order efflux back to plasma.
PFOS is not metabolised; elimination is urinary and faecal (biliary plus
unabsorbed). The gestational models add a placenta and a physically
separate fetal circulation (fetal plasma, liver, rest of body) linked to
the mother only by bidirectional placental diffusion, plus exchange with
amniotic fluid. The lactational models replace the fetus with a milk
compartment and a nursing pup/neonate submodel fed by suckled milk.

**Model time is not time since dose.** Every maternal, fetal and
neonatal volume and blood flow in the gestational and lactational models
is an explicit function of gestational or postnatal age, so time is
measured from GD0 / conception (gestation: GD = time/24 days, GA =
time/168 weeks) or from delivery (lactation: PND = time/24 days). Body
weight enters as the covariate `WT`, interpreted as the pre-pregnancy
maternal weight BW0 (gestation) or the delivery weight (lactation) that
anchors the growth equations.

## Provenance

The paper prints the chemical parameters (Table 2) and describes the
growth equations in prose, but the polynomial coefficients live in
Supplemental Tables S1-S6 and the executable model is the authors' openly
published mrgsolve code (<https://github.com/KSUICCM/PFOS-Ges-Lac>, also
deposited as the EHP supplemental zip). The maintainers took the model
structure, the physiological growth equations and the **as-run**
calibrated parameters directly from that code and its fit objects
(`GFit_R.rds`, `LFit_R.rds`, `GFit_H.rds`, `LFit_H.rds`). Where a fit
object and the printed Table 2 disagree, the code is treated as the
source of record because it is what generated the published simulations
(see Assumptions and deviations).

```{r load-models}
rat_pre  <- rxode2::rxode2(readModelDb("Chou_2021_pfos_rat_prepregnant_pbpk"))
rat_gest <- rxode2::rxode2(readModelDb("Chou_2021_pfos_rat_gestational_pbpk"))
rat_lact <- rxode2::rxode2(readModelDb("Chou_2021_pfos_rat_lactational_pbpk"))
hum_pre  <- rxode2::rxode2(readModelDb("Chou_2021_pfos_human_prepregnant_pbpk"))
hum_gest <- rxode2::rxode2(readModelDb("Chou_2021_pfos_human_gestational_pbpk"))
hum_lact <- rxode2::rxode2(readModelDb("Chou_2021_pfos_human_lactational_pbpk"))
c(rat_pre = length(rat_pre$state), rat_gest = length(rat_gest$state),
  rat_lact = length(rat_lact$state), hum_pre = length(hum_pre$state),
  hum_gest = length(hum_gest$state), hum_lact = length(hum_lact$state))
```

The models are stiff at early gestation and lactation, where the growth
equations make several fetal / neonatal volumes vanishingly small. The
default `liblsoda` solver with tightened tolerances is used throughout;
it is the most robust here (the single-threaded `lsoda` and `dop853`
fail on the rat gestational model).

```{r solver}
SOLVER <- list(method = "liblsoda", atol = 1e-10, rtol = 1e-8)
solve_model <- function(model, events, ...) {
  do.call(
    rxode2::rxSolve,
    c(list(object = model, events = events, returnType = "data.frame"), SOLVER, list(...))
  )
}
```

# Population

**Rat (Chou and Lin 2021 Table 1).** Sprague-Dawley rats from four
toxicokinetic studies were dosed by oral gavage at 0.1-10 mg/kg per day
before, during and after gestation; maternal plasma and liver, and
fetal/pup plasma and liver, were sampled from GD7 through PND21. The
reference pre-pregnant weight is 0.185 kg, the gestational dosing weight
0.225 kg, the lactational dosing weight 0.242 kg, and the litter is
eight fetuses/pups.

**Human (Chou and Lin 2021 Table 1).** More than twenty human
biomonitoring cohorts contributed maternal plasma, cord blood, placenta,
fetal tissue and breast-milk PFOS concentrations across gestation and
lactation. Exposure is chronic dietary intake; population-specific daily
intakes of 0.19-4.4 ng/kg per day (Loccisano et al. 2013) were used. The
reference pre-pregnant weight is 60 kg and the delivery weight excluding
the fetus is 68.05 kg.

# Source trace

| Component | Source |
| --- | --- |
| Oral uptake (stomach, small intestine) | Eqs 1-3 |
| Urinary and faecal elimination | Eqs 4-5 |
| Kidney: filtration, MM reabsorption, diffusion, efflux | Eqs 6-12 |
| Placental and amniotic-fluid transfer | Eqs 13-20 |
| Milk transfer and suckling | Eqs 21-23 |
| Chemical parameters, rat and human, per life stage | Table 2, Tables S1-S6 |
| Rat gestational calibrated set | `GFit_R.rds` |
| Rat lactational calibrated set | `LFit_R.rds` |
| Human gestational calibrated set | `GFit_H.rds` |
| Human lactational calibrated set | `LFit_H.rds` |
| Physiological growth equations | Kapraun et al. 2019; Loccisano et al. 2012, 2013; Yang et al. 2019 |

# Mass-balance gate

Each model conserves mass: the administered PFOS equals the sum of drug
in every tissue and excretion compartment (the AUC accumulators are not
mass and are excluded). This is a numerical check on the ODE encoding, so
a tight bound is used.

```{r mass-balance}
mass_states <- function(model) {
  st <- model$state
  st[!grepl("^auc_", st)]
}
mass_balance_err <- function(model, events, dose_mg, drop = character(0)) {
  s <- solve_model(model, events)
  last <- s[nrow(s), ]
  states <- setdiff(mass_states(model), drop)
  tot <- sum(vapply(states, function(v) last[[v]], numeric(1)))
  (dose_mg - tot) / dose_mg
}

# Rat prepregnant: single 1 mg/kg oral dose
ev_rp <- rxode2::et(amt = 0.185, cmt = "stomach") %>% rxode2::et(seq(0, 720, by = 24))
ev_rp$WT <- 0.185
err_rp <- mass_balance_err(rat_pre, ev_rp, 0.185)

# Rat gestational: 1 mg/kg/day GD2-GD20 (Thibodeaux 2003), dose ref 0.225 kg
ev_rg <- rxode2::et(amt = 0.225, cmt = "stomach", time = 48, ii = 24, addl = 18) %>%
  rxode2::et(seq(0, 504, by = 12))
ev_rg$WT <- 0.185
err_rg <- mass_balance_err(rat_gest, ev_rg, 19 * 0.225)

# Rat lactational: 1 mg/kg/day PND0-PND21, dose ref 0.242 kg. Pup urine is
# returned to the dam gut (coprophagy), so `urine_pup` is a throughput tracker
# whose mass has already re-entered the dam system; it is excluded from the
# closed-system total to avoid double counting.
ev_rl <- rxode2::et(amt = 0.242, cmt = "stomach", time = 0, ii = 24, addl = 21) %>%
  rxode2::et(seq(0, 504, by = 12))
ev_rl$WT <- 0.247
err_rl <- mass_balance_err(rat_lact, ev_rl, 22 * 0.242, drop = "urine_pup")

# Human prepregnant: single dietary dose
ev_hp <- rxode2::et(amt = 1e-4, cmt = "stomach") %>% rxode2::et(seq(0, 8760, by = 168))
ev_hp$WT <- 60
err_hp <- mass_balance_err(hum_pre, ev_hp, 1e-4)

# Human gestational: 2 ng/kg/day for 40 weeks
ev_hg <- rxode2::et(amt = 1.2e-4, cmt = "stomach", time = 0, ii = 24, addl = 279) %>%
  rxode2::et(seq(0, 280 * 24, by = 168))
ev_hg$WT <- 60
err_hg <- mass_balance_err(hum_gest, ev_hg, 280 * 1.2e-4)

# Human lactational: 2 ng/kg/day for 6 months. The output grid runs past the
# last dose (day 179) so every administered dose is inside the closed system.
ev_hl <- rxode2::et(amt = 1.36e-4, cmt = "stomach", time = 0, ii = 24, addl = 179) %>%
  rxode2::et(seq(0, 180 * 24, by = 24))
ev_hl$WT <- 68.05
err_hl <- mass_balance_err(hum_lact, ev_hl, 180 * 1.36e-4)

mb <- data.frame(
  model = c("rat prepregnant", "rat gestational", "rat lactational",
            "human prepregnant", "human gestational", "human lactational"),
  rel_err = c(err_rp, err_rg, err_rl, err_hp, err_hg, err_hl)
)
knitr::kable(mb, digits = 12, caption = "Relative mass-balance error, all six models.")
stopifnot(all(abs(mb$rel_err) < 1e-6))
```

# Rat gestational model versus observed maternal plasma

The rat gestational model is checked against the maternal-plasma PFOS
concentrations of Thibodeaux et al. (2003), in which dams were dosed
1-10 mg/kg per day from GD2 through GD20 (Chou and Lin 2021 Figure 3,
calibration data in the authors' repository). We simulate each dose and
compare the predicted maternal plasma at GD7, GD14 and GD21 with the
observed values.

```{r rat-gof}
# Observed maternal plasma (mg/L) from Thibodeaux 2003 (Study 1), as used in
# the authors' calibration (Data_R.csv). Rows: GD7, GD14, GD21.
obs_rat <- tibble::tribble(
  ~Dose, ~GD, ~Conc,
  1,  7, 14.82, 1, 14, 27.53, 1, 21, 20.65,
  2,  7, 25.94, 2, 14, 53.48, 2, 21, 45.53,
  3,  7, 53.47, 3, 14, 94.25, 3, 21, 70.42,
  5,  7, 76.25, 5, 14, 146.67, 5, 21, 81.54,
  10, 7, 154.82, 10, 14, 285.29, 10, 21, 188.59
)

pred_rat_gest <- function(dose_mgkg) {
  ev <- rxode2::et(amt = dose_mgkg * 0.225, cmt = "stomach", time = 48, ii = 24, addl = 18) %>%
    rxode2::et(c(7, 14, 21) * 24)
  ev$WT <- 0.185
  s <- solve_model(rat_gest, ev)
  s <- s[s$time %in% (c(7, 14, 21) * 24), ]
  data.frame(Dose = dose_mgkg, GD = s$time / 24, pred = s$Cc)
}
pred_rat <- do.call(rbind, lapply(c(1, 2, 3, 5, 10), pred_rat_gest))

gof_rat <- merge(obs_rat, pred_rat, by = c("Dose", "GD"))
gof_rat$ratio <- gof_rat$pred / gof_rat$Conc
knitr::kable(gof_rat, digits = 2,
  caption = "Rat gestational: predicted vs observed maternal plasma PFOS (mg/L), Thibodeaux 2003.")

within2 <- mean(gof_rat$ratio > 0.5 & gof_rat$ratio < 2)
cat(sprintf("Within 2-fold: %.0f%% (%d of %d points); median pred/obs = %.2f\n",
            100 * within2, sum(gof_rat$ratio > 0.5 & gof_rat$ratio < 2),
            nrow(gof_rat), stats::median(gof_rat$ratio)))
# The paper's acceptance criterion is prediction within a factor of 2.
stopifnot(within2 >= 0.8)
```

```{r rat-gof-plot, fig.cap = "Rat gestational maternal-plasma PFOS: predicted (line of identity) versus observed (Thibodeaux 2003). Dashed lines bound the 2-fold envelope."}
ggplot2::ggplot(gof_rat, ggplot2::aes(Conc, pred, colour = factor(Dose))) +
  ggplot2::geom_abline(slope = 1, intercept = 0) +
  ggplot2::geom_abline(slope = 2, intercept = 0, linetype = 2) +
  ggplot2::geom_abline(slope = 0.5, intercept = 0, linetype = 2) +
  ggplot2::geom_point(size = 2) +
  ggplot2::scale_x_log10() + ggplot2::scale_y_log10() +
  ggplot2::labs(x = "Observed maternal plasma PFOS (mg/L)",
                y = "Predicted (mg/L)", colour = "Dose (mg/kg/d)") +
  ggplot2::theme_bw()
```

# Rat lactational model: dam, milk and pups

The lactational dam is dosed 1 mg/kg per day PND0-PND21; PFOS reaches the
nursing pups only through milk. The dam plasma stays near its gestational
end value while pup plasma and liver accumulate over lactation.

```{r rat-lact}
ev_rl2 <- rxode2::et(amt = 0.242, cmt = "stomach", time = 0, ii = 24, addl = 21) %>%
  rxode2::et(seq(0, 504, by = 12))
ev_rl2$WT <- 0.247
s_rl <- solve_model(rat_lact, ev_rl2)
end_rl <- s_rl[nrow(s_rl), ]
cat(sprintf("PND21: dam plasma %.2f, milk %.4f, pup plasma %.2f, pup liver %.2f mg/L\n",
            end_rl$Cc, end_rl$Cmilk, end_rl$Cc_pup, end_rl$Cliver_pup))
# Pups accumulate PFOS from milk; pup liver exceeds pup plasma (PL_pup > 1).
stopifnot(end_rl$Cc_pup > 0, end_rl$Cliver_pup > end_rl$Cc_pup)
```

```{r rat-lact-plot, fig.cap = "Rat lactational PFOS: dam plasma and nursing-pup plasma over PND0-21 at 1 mg/kg/day."}
s_rl %>%
  dplyr::transmute(PND = time / 24, `Dam plasma` = Cc, `Pup plasma` = Cc_pup) %>%
  tidyr::pivot_longer(-PND) %>%
  ggplot2::ggplot(ggplot2::aes(PND, value, colour = name)) +
  ggplot2::geom_line() +
  ggplot2::labs(x = "Postnatal day", y = "PFOS (mg/L)", colour = NULL) +
  ggplot2::theme_bw()
```

# Human gestational model: cord-to-maternal ratio

The central quantitative feature of the human gestational model is
transplacental transfer. The paper's calibration cohorts report cord
blood (fetal plasma) at roughly 0.3-0.4 times maternal plasma at term
(e.g. Inoue 2004 CB/MP = 2.9/8.9 = 0.33; Fei 2007 = 11/29.2 = 0.38;
Kato 2014 = 3.5/8.5 = 0.41). We reproduce a chronic dietary exposure over
40 weeks and read the term ratio, which is independent of the absolute
intake.

```{r human-gest-ratio}
ev_hg2 <- rxode2::et(amt = 1.2e-4, cmt = "stomach", time = 0, ii = 24, addl = 279) %>%
  rxode2::et(seq(0, 280 * 24, by = 168))
ev_hg2$WT <- 60
s_hg <- solve_model(hum_gest, ev_hg2)
term <- s_hg[nrow(s_hg), ]
cbmp <- term$Cc_fet / term$Cc
cat(sprintf("Term (GA 40 wk) cord:maternal plasma ratio = %.3f\n", cbmp))
# Observed CB/MP across the calibration cohorts is ~0.33-0.41.
stopifnot(cbmp > 0.2, cbmp < 0.6)
```

```{r human-gest-plot, fig.cap = "Human gestational PFOS: maternal and fetal (cord) plasma over 40 weeks of constant dietary intake."}
s_hg %>%
  dplyr::transmute(GA = time / 168, `Maternal plasma` = Cc, `Fetal plasma` = Cc_fet) %>%
  ggplot2::ggplot(ggplot2::aes(GA)) +
  ggplot2::geom_line(ggplot2::aes(y = `Maternal plasma`, colour = "Maternal")) +
  ggplot2::geom_line(ggplot2::aes(y = `Fetal plasma`, colour = "Fetal")) +
  ggplot2::labs(x = "Gestational age (weeks)", y = "PFOS (mg/L)", colour = NULL) +
  ggplot2::theme_bw()
```

# Human pre-pregnant burn-in

The pre-pregnant model is run with constant dietary intake to steady
state; the paper simulates birth to 30 years to seed the gestational
run. Chronic dietary intake is represented as a continuous zero-order
input into the stomach (smoother, and more faithful to a diet than daily
boluses). A 20-year run reaches steady state.

```{r human-prepreg}
years <- 20
Tend  <- years * 365 * 24
rate  <- 1.2e-4 / 24 # 1.2 ng/kg/day * 60 kg as a per-hour intake
ev_hp2 <- rxode2::et(amt = rate * Tend, rate = rate, cmt = "stomach") %>%
  rxode2::et(seq(0, Tend, by = 24 * 90))
ev_hp2$WT <- 60
s_hp <- solve_model(hum_pre, ev_hp2)
last2 <- tail(s_hp$Cc, 2)
cat(sprintf("Plasma at year %d = %.4f mg/L; change over last quarter = %.2f%%\n",
            years, tail(s_hp$Cc, 1), 100 * (last2[2] - last2[1]) / last2[2]))
# Approaching steady state: the last quarterly change is small, and PFOS has
# accumulated relative to early exposure.
stopifnot(abs((last2[2] - last2[1]) / last2[2]) < 0.02, tail(s_hp$Cc, 1) > s_hp$Cc[2])
```

# Human lactational model: milk and neonate

The lactating mother transfers PFOS to the breastfed neonate through
milk; milk feeding stops after month 6 in the model.

```{r human-lact}
ev_hl2 <- rxode2::et(amt = 1.36e-4, cmt = "stomach", time = 0, ii = 24, addl = 179) %>%
  rxode2::et(seq(0, 180 * 24, by = 168))
ev_hl2$WT <- 68.05
s_hl <- solve_model(hum_lact, ev_hl2)
end_hl <- s_hl[nrow(s_hl), ]
cat(sprintf("Month 6: maternal plasma %.5f, milk %.6f, neonate plasma %.5f mg/L; milk:plasma = %.3f\n",
            end_hl$Cc, end_hl$Cmilk, end_hl$Cc_neo, end_hl$Cmilk / end_hl$Cc))
# Milk:plasma partitioning is of order PMilkP (~0.01); the neonate takes up PFOS.
stopifnot(end_hl$Cc_neo > 0, end_hl$Cmilk > 0)
```

# Assumptions and deviations

- **Source of record.** Structure, growth equations and calibrated
  parameters were taken from the authors' openly published mrgsolve code
  and fit objects, not the paper text alone, because the polynomial
  growth coefficients are in Supplemental Tables S1-S6 and the executable
  model is the code. The calibrated values applied are the
  sensitivity-selected subsets stored in `GFit_R/LFit_R/GFit_H/LFit_H`;
  all other parameters keep the code's `$PARAM` defaults.
- **Table 2 versus code Km labelling.** In the rat and human gestational
  fits the calibrated Michaelis constant is stored under the code
  parameter named `Km_apical` (rat 19.89 mg/L; human 247.8 mg/L), whereas
  Table 2 labels the calibrated value `Km_baso`. The as-run code is
  reproduced faithfully (calibrated `Km_apical`, default `Km_baso`), which
  is what produced the published simulations; the labelling discrepancy is
  noted here rather than silently reconciled.
- **Deterministic.** The published Monte Carlo analysis draws
  physiological and chemical parameters from external distributions
  (typically 20-30% CV, truncated) around these point estimates; the
  model files carry the point estimates only, with no random effects. A
  `propSd` residual-error placeholder is fixed and carries no information
  from the paper.
- **Body weight.** `WT` is the single covariate. In the gestational
  models it is the pre-pregnancy weight BW0 that anchors the growth
  equations (gestational weight BW_P is computed internally); the authors
  dose against a separate gestational/lactational reference weight when
  converting a mg/kg dose to a mg amount, which the simulations above
  reproduce. The published human pre-pregnant model holds body weight
  constant rather than using the age-dependent equation described in the
  text.
- **Litter.** The rat litter is modelled as a single fetal/pup
  compartment scaled by N = 8, and pup urine is returned to the dam gut
  (coprophagy), both as in the authors' code.
- **Specimen metadata.** Amniotic fluid and the AUC accumulator
  compartments have no matching entry in the package specimen vocabulary
  and are recorded as `not applicable`; placental, fetal, pup and
  neonatal solid compartments are recorded as `tissue`.
- **Absolute human concentrations** depend on the assumed dietary intake,
  which is uncertain; the human checks above therefore assert on the
  intake-independent cord-to-maternal ratio and on approach to steady
  state rather than on absolute plasma levels.
