---
title: "Motesanib tumor size and overall survival in NSCLC (Claret 2014)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Motesanib tumor size and overall survival in NSCLC (Claret 2014)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

This paper contributes two models to the library.

```{r load-models}
tso <- rxode2::rxode(readModelDb("Claret_2014_motesanib_tumorsize_OS"))
nev <- rxode2::rxode(readModelDb("Claret_2014_motesanib_OS_nonevaluable"))
```

* Citation: `r tso$reference`
* Article: <https://doi.org/10.1038/clpt.2014.11>

**`Claret_2014_motesanib_tumorsize_OS`** -- the joint tumor-size (TS) /
overall-survival (OS) model for the 934 TS-*evaluable* patients.

**`Claret_2014_motesanib_OS_nonevaluable`** -- the separate OS model for the
patients whose tumor-size response could not be predicted, which the authors
used to correct their trial simulations for the nonevaluable fraction.

A third model is discussed in the paper but is **not** extractable from it: the
"historical" first-line NSCLC OS model of Wang et al., which Claret 2014
assesses and then improves upon. Its parameter estimates live in that earlier
publication and are not reprinted here, so no model file is created for it.

# Population

MONET1 randomised 1,090 patients with advanced or recurrent nonsquamous NSCLC
to carboplatin/paclitaxel (CP) plus motesanib 125 mg once daily or CP plus
placebo, first line. Of these, 934 (86%) were TS-evaluable --
453 of 541 on motesanib and 481 of 549 on placebo -- with a median of four TS
measurements per patient (range 2-27). The remaining 156 (14%) were
nonevaluable: 115 had no postbaseline tumor assessment and 41 had only
nontarget lesions. 227 enrolled patients (21%) were Asian; the 219 Asian
patients used for the virtual phase III simulations included 21 nonevaluable
patients.

The same information is available programmatically from each model's
`population` metadata:

```{r population}
str(readModelDb("Claret_2014_motesanib_tumorsize_OS")()$population[
  c("n_subjects", "disease_state", "race_ethnicity")
])
```

# Source trace

Every `ini()` value carries an in-file comment naming its source location. They
are collected here for review.

| Equation / parameter | Value | Source location |
|---|---|---|
| `TS(t) = BASE * exp(KL*t - (KDE/lambda)*(1 - exp(-lambda*t)))` | n/a | Methods, "TS model" (p. 450, left column) |
| `d/dt(tumor_size) = tumor_size * (kge - kdrug*exp(-lambda*t))` | n/a | Derivative of the Methods TS equation (see "Structural checks" below) |
| `TSratio = TS(week8) / BASE` | n/a | Methods, "Two metrics to assess TS response", metric 1 |
| `TTG = (log(KDE) - log(KL)) / lambda` | n/a | Methods, metric 2 |
| `log(TTG*) = log(TTG + 100)` | n/a | Methods, shifted-log definition |
| `lrbase` (BASE) | 90 mm | Supplementary Table S1 (RSE 2.37%) |
| `lkge` (KL) | 0.00107 /day | Supplementary Table S1 (RSE 7.70%) |
| `lkdrug` (KDE) | 0.0103 /day | Supplementary Table S1, estimate read off the *lambda* row (RSE 5.16% from the KDE row); see "Supplementary Table S1 transposes KDE and lambda" and Errata item 1 |
| `llambda` (lambda) | 0.0193 /day | Supplementary Table S1, estimate read off the *KDE* row (RSE 3.72% from the lambda row); see "Supplementary Table S1 transposes KDE and lambda" and Errata item 1 |
| `etalrbase` | SD 0.683 -> var 0.4665 | Supplementary Table S1, "SD" column |
| `etalkge` | SD 1.18 -> var 1.3924 | Supplementary Table S1, "SD" column |
| `etalkdrug` | SD 0.679 -> var 0.4610 | Supplementary Table S1, "SD" column |
| `etallambda` | SD 0.797 -> var 0.6352 | Supplementary Table S1, "SD" column |
| `addSd` (Sigma) | 7.849 mm | Supplementary Table S1 (RSE 1.18%) |
| `mu_os_int_logdays` | 1.079 | Table 3, Intercept (SE 0.436) |
| `e_tsb_mu_os` | -0.00197 /mm | Table 3, Baseline TS (SE 0.00035) |
| `e_ttg_mu_os` | 0.97695 | Table 3, Log(TTG*(day)) (SE 0.0794) |
| `e_race_asian_mu_os` | 0.319 | Table 3, Asian ethnicity (SE 0.069) |
| `e_smoke_ever_mu_os` | -0.264 | Table 3, Smoking history (SE 0.063) |
| `sigma_os_log` | -0.24 | Table 3, Log(scale) (SE 0.0282) |
| Nonevaluable `mu_os_int_logdays` | 0.681 | Supplementary Table S2, Intercept (SE 0.881) |
| Nonevaluable `e_alb_mu_os` | 0.108 /(g/L) | Supplementary Table S2, Baseline albumin (SE 0.0237) |
| Nonevaluable `sigma_os_log` | 0.502 | Supplementary Table S2, Log(scale) (SE 0.0702) |
| TTG time unit (days, not the Methods' "weeks") | n/a | Table 3 row heading `Log(TTG*(day))` + footnote c; see Errata item 11 |
| Nonevaluable model VPC (validation target) | n/a | Supplementary Figure S3 |
| TS goodness-of-fit (axes mislabelled "cm") | n/a | Supplementary Figure S1; see Errata item 7 |

Parameter covariance matrices for the two OS models are in Supplementary Tables
S3 and S4 respectively. They are not encoded in the model files -- an nlmixr2
model object carries point estimates, not the estimation covariance -- but they
are what the paper used to propagate "model uncertainty at the replicate level"
through its trial simulations. Each model file's `reference` field names the
supplementary table to read them from.

# Mechanism in one paragraph

Tumor size grows exponentially at rate `kge` (the paper's KL) and is suppressed by
a drug term that starts at `kdrug` (KDE) and decays exponentially at rate
`lambda`, representing acquired resistance and dose reductions. Net growth is
therefore negative early and turns positive once `kdrug * exp(-lambda*t)` falls
below `kge`. The instant that happens is the tumor-size nadir, and the time to it
is **TTG** -- the metric this paper proposes as the driver of survival benefit.
Because TTG rewards *duration* of effect and not just depth of shrinkage, it
credits a cytostatic antiangiogenic such as motesanib in a way that the week-8
shrinkage ratio (TSratio) does not. TTG then enters a log-normal accelerated
failure time (AFT) survival model as `log(TTG + 100)`, alongside baseline tumor
size, Asian ethnicity and smoking history.

# Structural checks

These are exact algebraic identities of the encoded model, so they are asserted
tightly. They confirm that the ODE encoding, the TTG formula and the AFT
survivor function are all faithful to the printed equations.

## The ODE reproduces the published closed form

The model file encodes the TS equation as an ODE. Differentiating the paper's
closed form gives `d(TS)/dt = TS * (KL - KDE*exp(-lambda*t))`, so the two must
agree to solver tolerance.

```{r check-ode}
# KDE and lambda take the CORRECTED (de-transposed) assignment used by the
# model file -- see "Supplementary Table S1 transposes KDE and lambda" below.
BASE <- 90; KL <- 0.00107; KDE <- 0.0103; LAM <- 0.0193
tgrid <- seq(0, 1100, by = 5)
closed_form <- BASE * exp(KL * tgrid - (KDE / LAM) * (1 - exp(-LAM * tgrid)))

ev_typ <- data.frame(
  id = 1L, time = tgrid, cmt = "tumor_size", evid = 0L, amt = NA_real_,
  RACE_ASIAN = 0, SMOKE_NEVER = 0
)
sim_typ <- as.data.frame(
  rxode2::rxSolve(rxode2::zeroRe(tso), ev_typ, returnType = "data.frame")
)

max_rel_err <- max(abs(sim_typ$tumor_size - closed_form) / closed_form)
max_rel_err
stopifnot(max_rel_err < 1e-4)
```

## TTG is the time to nadir

`TTG = (log(KDE) - log(KL)) / lambda` should land exactly on the minimum of the
closed-form curve.

```{r check-ttg}
fine <- seq(0, 1100, by = 0.05)
nadir_numeric <- fine[which.min(
  BASE * exp(KL * fine - (KDE / LAM) * (1 - exp(-LAM * fine)))
)]
ttg_model <- unique(sim_typ$ttg)

c(ttg_formula = ttg_model, nadir_numeric = nadir_numeric)
stopifnot(abs(ttg_model - nadir_numeric) < 0.1)
```

## The AFT survivor function is internally consistent

Two identities: the cumulative hazard must be `-log(S(t))`, and the model's
`median_os` output must be the time at which `S(t) = 0.5`.

```{r check-survivor}
stopifnot(max(abs(sim_typ$sur - exp(-sim_typ$cumhazard))) < 1e-12)

med  <- unique(sim_typ$median_os)
zmed <- (log(med) - unique(sim_typ$mu_os_logdays)) / unique(sim_typ$sigma_os)
s_at_median <- 1 - pnorm(zmed)

c(median_os_days = med, S_at_median = s_at_median)
stopifnot(abs(s_at_median - 0.5) < 1e-8)
```

# Tumor-size trajectories

```{r cohort}
# rxSetSeed fixes the eta draw within an rxode2 version; the summaries below
# are reported, not asserted on, precisely because the extreme quantiles of a
# random cohort are not reproducible across rxode2 builds.
rxode2::rxSetSeed(20140311)
set.seed(20140311)
n_sub <- 200L

obs_times <- seq(0, 1100, by = 14)
cohort <- tidyr::expand_grid(id = seq_len(n_sub), time = obs_times) |>
  dplyr::mutate(cmt = "tumor_size", evid = 0L, amt = NA_real_,
                RACE_ASIAN = 0, SMOKE_NEVER = 0)

sim <- as.data.frame(rxode2::rxSolve(tso, cohort, returnType = "data.frame"))
```

```{r fig-ts}
band <- sim |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    Q05 = quantile(tumor_size, 0.05),
    Q50 = quantile(tumor_size, 0.50),
    Q95 = quantile(tumor_size, 0.95),
    .groups = "drop"
  )

ggplot(band, aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(linewidth = 0.8, colour = "steelblue") +
  geom_line(
    data = data.frame(time = tgrid, Q50 = closed_form),
    linetype = "dashed", colour = "black"
  ) +
  geom_vline(xintercept = ttg_model, linetype = "dotted") +
  labs(
    x = "Time (days)", y = "Tumor size (mm)",
    title = "Simplified TGI model: median and 5th-95th percentile band",
    caption = paste0(
      "Dashed = typical-value (zeroRe) trajectory; dotted vertical line = ",
      "typical-value TTG (", round(ttg_model), " days), the model-predicted nadir."
    )
  ) +
  theme_bw()
```

The characteristic shape is reproduced: shrinkage while the drug term dominates,
a nadir at TTG, then regrowth as resistance takes over.

# Derived TS metrics versus the published distributions

Claret 2014 reports the distributions of both TS metrics computed from
individual empirical-Bayes (EBE) estimates. These are the model's primary
validation targets, and reproducing them is what identified the transposition
of KDE and lambda in Supplementary Table S1 (see the dedicated section below,
and Errata item 1). With that correction applied, both metrics fall where the
paper puts them.

```{r ts-metrics}
per_subject <- sim |> dplyr::filter(time == 0)
ts_w8   <- sim |> dplyr::filter(time == 56)  |> dplyr::pull(tumor_size)
ts_base <- sim |> dplyr::filter(time == 0)   |> dplyr::pull(tumor_size)
tsratio <- ts_w8 / ts_base

metric_tbl <- tibble::tibble(
  Metric = c("TTG (days)", "TSratio (week 8 / baseline)"),
  `Simulated median` = c(
    round(median(per_subject$ttg), 1),
    round(median(tsratio), 2)
  ),
  `Simulated 2.5-97.5%` = c(
    paste(round(quantile(per_subject$ttg, c(0.025, 0.975)), 0), collapse = " to "),
    paste(round(quantile(tsratio, c(0.025, 0.975)), 2), collapse = " to ")
  ),
  `Published, motesanib arm` = c("125 (28 to 399)", "0.69 (0.24 to 0.99)"),
  `Published, placebo arm`   = c("93 (-6 to 312)",  "0.77 (0.27 to 1.25)")
)
knitr::kable(metric_tbl)
```

The tightest available check is on the **typical value** -- all etas zero --
because the TS model is a single pooled fit with no treatment-arm term. Its
typical subject is therefore the pooled-cohort central subject, and its metrics
must land *between* the two published per-arm medians. That statement involves
no random draw at all, so it is asserted exactly.

```{r ts-metrics-typical}
ttg_typ     <- (log(KDE) - log(KL)) / LAM
tsratio_typ <- exp(KL * 56 - (KDE / LAM) * (1 - exp(-LAM * 56)))

c(TTG_typical = round(ttg_typ, 1), TSratio_typical = round(tsratio_typ, 3))

stopifnot(
  # Published per-arm medians: TTG 93 (placebo) and 125 (motesanib) days.
  ttg_typ > 93, ttg_typ < 125,
  # Published per-arm medians: TSratio 0.69 (motesanib) and 0.77 (placebo).
  tsratio_typ > 0.69, tsratio_typ < 0.77
)
```

Both typical values sit inside the published arm interval, which is what a
pooled no-arm-term fit is supposed to do.

The cohort medians agree too, but they are asserted much more loosely: TTG is a
ratio of log-normal quantities and is heavy-tailed, so a 200-subject median
carries real sampling noise and its upper quantiles are not reproducible across
rxode2 builds. The upper tail is deliberately not asserted on -- a forward draw
from a diagonal prior is also genuinely wider than a distribution of shrunken
EBEs, so the published 97.5th percentiles are expected to be *narrower* than
the simulated ones.

```{r ts-metrics-assert}
stopifnot(
  # Centre: within 40% of the published pooled median (~108 days). A
  # mis-transcribed rate constant moves this by 2- to 3-fold.
  abs(median(per_subject$ttg) / 108 - 1) < 0.4,
  # TTG is positive for the large majority of subjects, as reported (the
  # published placebo 2.5th percentile is -6 days, i.e. ~2.5% negative).
  # The closed-form probability is pnorm(log(KDE/KL)/sqrt(varKDE+varKL)) =
  # 0.952, so at n = 200 the binomial SE is 0.015; the bound is set ~7 SE
  # below it rather than the ~3 SE that 0.9 would give, because the eta draw
  # is not reproducible across rxode2 versions. This is a sanity check on the
  # sign of the drug effect, not a discriminator for the Table S1 erratum --
  # the printed reading would give a HIGHER positive fraction (0.983).
  mean(per_subject$ttg > 0) > 0.85,
  # Week-8 TSratio: net shrinkage, centred on the published 0.69-0.77.
  median(tsratio) < 1,
  abs(median(tsratio) / 0.73 - 1) < 0.25
)
```

## How often is `log(TTG + 100)` undefined?

TTG is negative whenever KL > KDE, and the published shifted log is undefined
for TTG <= -100 days. No guard is added because none is published. The exact
probability under the Supplementary Table S1 IIV is available in closed form --
`TTG = N / lambda` with `N ~ Normal(log(KDE) - log(KL), sqrt(var_KDE +
var_KL))` and `lambda` log-normal -- so it is integrated rather than estimated
by Monte Carlo:

```{r ttg-undefined}
mu_N    <- log(KDE) - log(KL)          # KDE/KL with the corrected assignment
sd_N    <- sqrt(0.679^2 + 1.18^2)      # SDs stay on their printed Table S1 rows
mu_lam  <- log(LAM)
sd_lam  <- 0.797

p_undefined <- integrate(
  function(lam) pnorm((-100 * lam - mu_N) / sd_N) * dlnorm(lam, mu_lam, sd_lam),
  lower = 0, upper = Inf
)$value

round(p_undefined, 5)
stopifnot(p_undefined < 0.02)
```

About `r sprintf("%.1f%%", 100 * p_undefined)` of subjects drawn from the
published IIV fall outside the domain of `log(TTG + 100)` and return non-finite
survival quantities. Filter them explicitly in downstream work; the observed
MONET1 TTG range (-6 to 399 days) is comfortably inside the domain.

# The OS model: covariate effects

On the AFT log-time scale every covariate is a multiplicative shift of median
survival, so the following are exact identities that pin each Table 3 estimate.

```{r os-covariates}
os_grid <- expand.grid(RACE_ASIAN = c(0, 1), SMOKE_NEVER = c(0, 1)) |>
  dplyr::mutate(id = dplyr::row_number())

os_ev <- os_grid |>
  tidyr::expand_grid(time = c(0, 365)) |>
  dplyr::mutate(cmt = "tumor_size", evid = 0L, amt = NA_real_)

# Select only the outputs before joining: rxSolve echoes the covariate
# columns back, so joining os_grid without this select would silently
# produce RACE_ASIAN.x / RACE_ASIAN.y and leave $RACE_ASIAN NULL.
os_sim <- as.data.frame(
  rxode2::rxSolve(rxode2::zeroRe(tso), os_ev, returnType = "data.frame")
) |>
  dplyr::distinct(id, .keep_all = TRUE) |>
  dplyr::select(id, median_os, mu_os_logdays) |>
  dplyr::left_join(os_grid, by = "id")

ref   <- os_sim$median_os[os_sim$RACE_ASIAN == 0 & os_sim$SMOKE_NEVER == 0]
asian <- os_sim$median_os[os_sim$RACE_ASIAN == 1 & os_sim$SMOKE_NEVER == 0]
never <- os_sim$median_os[os_sim$RACE_ASIAN == 0 & os_sim$SMOKE_NEVER == 1]

# Guard against an empty subset: `stopifnot()` passes vacuously on
# logical(0), so the length checks below are what make the ratio
# assertions meaningful.
stopifnot(length(ref) == 1L, length(asian) == 1L, length(never) == 1L)

knitr::kable(
  tibble::tibble(
    Contrast = c("Asian vs non-Asian", "Never- vs ever-smoker",
                 "Per +100 mm baseline TS"),
    `Model ratio` = c(asian / ref, never / ref, exp(-0.00197 * 100)),
    `Table 3 expectation` = c(exp(0.319), exp(0.264), exp(-0.197))
  ),
  digits = 4
)

stopifnot(
  abs(asian / ref - exp(0.319)) < 1e-8,
  abs(never / ref - exp(0.264)) < 1e-8
)
```

Both signs match the paper's prose: survival probability is higher in Asian
patients and lower in former and current smokers.

# Figure 1: survival by TTG

Figure 1 of Claret 2014 splits the observed Kaplan-Meier curves by quartile of
TTG. The quartile boundaries themselves are not published, so rather than invent
midpoints the model analogue below is drawn at the TTG values the paper *does*
print: the per-arm median and 2.5th/97.5th percentiles of the empirical-Bayes
TTG distribution (Results, "TS model").

```{r fig-1}
mu_os <- function(ttg, tumsz = 90, asian = 0, ever = 1) {
  1.079 - 0.00197 * tumsz + 0.97695 * log(ttg + 100) +
    0.319 * asian - 0.264 * ever
}
sigma_os_aft <- exp(-0.24)

ttg_landmarks <- c(
  "2.5th pct, placebo (-6 d)"      = -6,
  "median, placebo (93 d)"         = 93,
  "median, motesanib (125 d)"      = 125,
  "97.5th pct, motesanib (399 d)"  = 399
)

km <- tidyr::expand_grid(
  landmark = factor(names(ttg_landmarks), levels = names(ttg_landmarks)),
  time     = seq(1, 1200, by = 5)
) |>
  dplyr::mutate(
    ttg = unname(ttg_landmarks[as.character(landmark)]),
    sur = 1 - pnorm((log(time) - mu_os(ttg)) / sigma_os_aft)
  )

ggplot(km, aes(time, sur, linetype = landmark)) +
  geom_line(linewidth = 0.7) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Time (days)", y = "Survival probability", linetype = NULL,
       title = "Model-predicted survival across the published TTG range",
       caption = paste(
         "Reference subject: baseline TS 90 mm, non-Asian, ever-smoker.",
         "Compare Figure 1 of Claret 2014."
       )) +
  theme_bw()
```

```{r fig-1-table}
land_tbl <- tibble::tibble(
  `Published TTG landmark`           = names(ttg_landmarks),
  `TTG (days)`                       = unname(ttg_landmarks),
  `Median OS, ever-smoker (days)`    = round(exp(mu_os(unname(ttg_landmarks)))),
  `Median OS, never-smoker (days)`   = round(exp(mu_os(unname(ttg_landmarks), ever = 0)))
)
knitr::kable(land_tbl)
```

Across the published TTG range the model's median survival for this reference
subject spans roughly 160 days (2.5th percentile, ever-smoker) to 1,070 days
(97.5th percentile, never-smoker). Figure 1's four observed quartile curves
cross S(t) = 0.5 in increasing order of TTG quartile, the first at roughly 200
days and the fourth somewhere between 600 and 800 days. No more precise
reading is quoted here: the second- and third-quartile curves cross each other
close to S(t) = 0.5, so their individual crossing times cannot be separated
reliably from the printed figure. What survives that uncertainty is that the
four medians are monotone in TTG quartile and lie inside the model's span,
which is the most that can be claimed.

Two reasons that comparison is qualitative only, and why nothing here is
asserted on it. First, a digitized curve is evidence about the printed figure,
not about the underlying patient data. Second, and more fundamentally, the two are not
like-for-like: Figure 1's curves pool *all* evaluable patients within a TTG
quartile, mixing ethnicity, smoking history and baseline tumor size, whereas the
model line is a single reference subject (90 mm, non-Asian, ever-smoker) at a
TTG *percentile* rather than a quartile midpoint. The paper does not publish the
quartile boundaries, so the two cannot be aligned.

The assertions below are therefore restricted to the two claims that are exact
properties of Table 3: survival increases monotonically with TTG -- the paper's
central claim -- and never-smokers outlive ever-smokers at equal TTG.

```{r fig-1-assert}
stopifnot(
  # Survival increases with TTG -- the paper's central claim.
  all(diff(exp(mu_os(unname(ttg_landmarks)))) > 0),
  # Never-smokers survive longer than ever-smokers at equal TTG (Table 3 sign).
  all(exp(mu_os(unname(ttg_landmarks), ever = 0)) >
        exp(mu_os(unname(ttg_landmarks))))
)
```

# Supplementary Table S1 transposes KDE and lambda

This is the one substantive correction this extraction makes to the published
material, so the evidence is laid out in full.

Supplementary Table S1 prints, in this row order:

| Parameter | Estimate | RSE, % | SD |
|---|---|---|---|
| KL (day^-1) | 0.00107 | 7.70 | 1.18 |
| KDE (day^-1) | **0.0193** | 5.16 | 0.679 |
| lambda (day^-1) | **0.0103** | 3.72 | 0.797 |
| BASE (mm) | 90 | 2.37 | 0.683 |
| Sigma (mm) | 7.849 | 1.18 | |

Read literally, that parameter set contradicts the paper's own Results in two
places at once, and by a wide margin. The model files therefore exchange the two
bolded **estimates**, using KDE = 0.0103 and lambda = 0.0193 per day. Both
numbers are taken verbatim from Table S1; only their row assignment changes, and
no value is tuned to a target.

## The decisive test does not involve the random effects

The TS model is a **single pooled fit with no treatment-arm term**. Its typical
subject -- every eta set to zero -- is therefore the central subject of the
pooled cohort, and its metrics must fall *between* the two published per-arm
medians. No covariance structure, eta correlation or empirical-Bayes shrinkage
argument can bear on this, because the typical value contains no etas at all.

```{r s1-decisive}
typ_metrics <- function(kde, lam, kl = 0.00107, t8 = 56) {
  c(TTG = (log(kde) - log(kl)) / lam,
    TSratio = exp(kl * t8 - (kde / lam) * (1 - exp(-lam * t8))))
}

decisive <- rbind(
  `as printed in Table S1` = typ_metrics(0.0193, 0.0103),
  `corrected (this model)` = typ_metrics(0.0103, 0.0193)
)

knitr::kable(
  tibble::tibble(
    Reading = rownames(decisive),
    `Typical TTG (days)` = round(decisive[, "TTG"], 1),
    `Published arm medians` = "93 to 125",
    `Typical TSratio` = round(decisive[, "TSratio"], 3),
    `Published arm medians ` = "0.69 to 0.77"
  )
)

stopifnot(
  # The printed reading falls outside BOTH published intervals.
  decisive["as printed in Table S1", "TTG"] > 125,
  decisive["as printed in Table S1", "TSratio"] < 0.69,
  # The corrected reading falls inside BOTH.
  decisive["corrected (this model)", "TTG"] > 93,
  decisive["corrected (this model)", "TTG"] < 125,
  decisive["corrected (this model)", "TSratio"] > 0.69,
  decisive["corrected (this model)", "TSratio"] < 0.77
)
```

The printed reading implies a typical patient who shrinks 53% by week 8 and does
not reach nadir until day 281. The paper reports median shrinkage of 23% and 31%
by arm and median nadir at days 93 and 125. That gap is far too large to be a
shrinkage artefact, and -- being eta-free -- it cannot be one.

## Two further published outputs agree independently

**The forward Monte Carlo over the Table S1 IIV.** This is a distributional
check rather than a point one, and it discriminates on the *lower* tail as well
as the centre. It is run in plain R rather than off the 200-subject rxode2
cohort so that it is reproducible across rxode2 builds.

```{r s1-montecarlo}
set.seed(20140311)
n_mc <- 2e5
mc_metrics <- function(kde, lam) {
  kl_i  <- 0.00107 * exp(rnorm(n_mc, 0, 1.18))
  kde_i <- kde     * exp(rnorm(n_mc, 0, 0.679))
  lam_i <- lam     * exp(rnorm(n_mc, 0, 0.797))
  ttg_i <- (log(kde_i) - log(kl_i)) / lam_i
  tsr_i <- exp(kl_i * 56 - (kde_i / lam_i) * (1 - exp(-lam_i * 56)))
  c(ttg_med = median(ttg_i),
    tsr_med = median(tsr_i),
    tsr_p025 = unname(quantile(tsr_i, 0.025)))
}

mc <- rbind(`as printed` = mc_metrics(0.0193, 0.0103),
            `corrected`  = mc_metrics(0.0103, 0.0193))

knitr::kable(
  tibble::tibble(
    Reading = c(rownames(mc), "published (pooled)"),
    `Median TTG (days)` = c(round(mc[, "ttg_med"]), 108),
    `Median TSratio` = c(round(mc[, "tsr_med"], 2), 0.73),
    `TSratio 2.5th pct` = c(round(mc[, "tsr_p025"], 3), 0.25)
  )
)

s1_ttg_md <- unname(mc["corrected", "ttg_med"])

stopifnot(
  # The corrected median TTG is within 40% of the published pooled ~108 days;
  # the printed one is more than double it.
  abs(mc["corrected", "ttg_med"] / 108 - 1) < 0.4,
  mc["as printed", "ttg_med"] > 2 * 108,
  # The published TSratio 2.5th percentile is 0.24-0.27 by arm. The corrected
  # reading lands in that neighbourhood; the printed one is ~5-fold below it.
  mc["corrected", "tsr_p025"] > 0.2,
  mc["as printed", "tsr_p025"] < 0.1
)
```

**The Table 5 simulation output, inverted through the Table 3 regression.**
Table 5 reports the paper's own simulated median OS for a virtual phase III
study in Asian patients: 594 days on CP + motesanib and 450 days on CP +
placebo. Those numbers are outputs of the *same* Table 3 regression, so the
regression can be inverted to recover the arm-level TTG the authors must have
been conditioning on. This uses only published values -- no digitization, no
reconstructed quartiles.

```{r ttg-scale}
invert_ttg <- function(median_os_days, tumsz = 90, asian = 1, ever = 1) {
  intercept <- 1.079 - 0.00197 * tumsz + 0.319 * asian - 0.264 * ever
  exp((log(median_os_days) - intercept) / 0.97695) - 100
}

tbl5_os   <- c(motesanib = 594, placebo = 450)
implied   <- invert_ttg(tbl5_os)
published <- c(motesanib = 125, placebo = 93)

knitr::kable(
  tibble::tibble(
    Arm = c("CP + motesanib", "CP + placebo"),
    `Table 5 simulated median OS (days)`  = unname(tbl5_os),
    `Implied TTG, ever-smoker (days)`     = round(unname(implied)),
    `Implied TTG, never-smoker (days)`    = round(unname(invert_ttg(tbl5_os, ever = 0))),
    `Published median TTG (days)`         = unname(published)
  )
)
```

Inverting Table 5 recovers arm-level TTGs in the same 90-160 day band as the
Results-text medians, and feeding the Results-text medians *forward* through the
Table 3 regression recovers Table 5's survival times -- 445 days computed
against 450 published on placebo. The printed Table S1 TTG of 281 days instead
implies a median OS of 864 days: 1.5 times Table 5's motesanib arm (594 days),
1.9 times its placebo arm (450 days), and 1.8 times the 479 days implied by the
published *pooled* TTG of about 108 days. That last ratio is the like-for-like
one -- 281 and 108 days are both pooled typical values -- and it is the
comparison asserted on below.

```{r ttg-scale-assert}
forward_os <- function(ttg, tumsz = 90, asian = 1, ever = 1) {
  exp(1.079 - 0.00197 * tumsz + 0.97695 * log(ttg + 100) +
        0.319 * asian - 0.264 * ever)
}

knitr::kable(
  tibble::tibble(
    `TTG source` = c("Results text, placebo (93 d)",
                     "Results text, motesanib (125 d)",
                     "Table S1 as printed, typical (281 d)",
                     "Table S1 corrected, typical (117 d)"),
    `Median OS via Table 3 (days)` = round(forward_os(c(93, 125, 280.8, 117.3))),
    `Table 5 published (days)` = c(450, 594, NA, NA)
  )
)

stopifnot(
  # Table 5 inverts to a longer TTG on motesanib than on placebo: the drug effect.
  implied["motesanib"] > implied["placebo"],
  # Table 5 and the Results-text medians agree within 50% on both arms.
  all(abs(implied / published - 1) < 0.5),
  # The published placebo TTG reproduces Table 5's placebo arm closely.
  abs(forward_os(93) / 450 - 1) < 0.1,
  # The printed Table S1 typical TTG overshoots even the BETTER of the two
  # published Table 5 arms by more than 40%.
  forward_os(280.8) > 1.4 * 594,
  # Like-for-like: both 280.8 d and 108 d are pooled/typical TTG values, so
  # this is the comparison the transposition actually distorts -- by >70%.
  forward_os(280.8) / forward_os(108) > 1.7,
  # The corrected typical TTG lands inside the published arm range.
  forward_os(117.3) > 450, forward_os(117.3) < 594
)
```

## Why a transposition, and why not something else

Four independent published quantities -- the Results-text TTG medians, the
Results-text TSratio medians and percentiles, the Table 3 regression and the
Table 5 simulation output -- all place the operative TTG scale near 100 days and
the week-8 TSratio near 0.73. Exchanging the two Table S1 estimates reproduces
every one of them; leaving them as printed misses each by 2- to 5-fold.

A transposition of two adjacent rows is also the most economical explanation.
Both candidate values already appear in the table, so nothing is invented, and a
swap moves *both* metrics coherently -- which a single mistyped digit would not,
since TTG and TSratio each depend on KDE and lambda together.

The competing explanation, that a forward draw from a diagonal prior is simply
the wrong comparator for a median of correlated empirical-Bayes estimates, is
ruled out by the decisive test above: it is an argument about the *distribution*
of etas, and the typical value has none.

One sub-choice is genuinely undetermined and is not load-bearing. The estimates
alone may be transposed (the reading used here: KDE keeps the printed KDE row's
RSE and SD), or the whole rows including labels may be swapped. The two differ
by about one day in median TTG (104 vs 105). The estimate-only reading is
preferred because it matches the published lower tails slightly better -- a
TSratio 2.5th percentile of 0.269 against 0.196, versus a published 0.24-0.27.

This has been recorded for author correspondence; until the authors confirm,
users who need to reproduce the paper's *printed* Table S1 exactly can override
the two values with `rxode2::ini()`.

# The nonevaluable-patient model

For the 14% of patients with no predictable TS metric, survival depends only on
baseline serum albumin.

```{r nonevaluable}
alb_grid <- data.frame(id = 1:5, time = 365, ALB = c(25, 30, 35, 40, 45))
nev_sim  <- as.data.frame(rxode2::rxSolve(nev, alb_grid, returnType = "data.frame"))

knitr::kable(
  nev_sim |>
    dplyr::select(ALB, median_os, sur) |>
    dplyr::rename(
      "Baseline albumin (g/L)" = ALB,
      "Median OS (days)" = median_os,
      "S(365 days)" = sur
    ),
  digits = c(0, 1, 3)
)

# Exact identity: +10 g/L albumin multiplies median OS by exp(10 * 0.108).
ratio_10 <- nev_sim$median_os[nev_sim$ALB == 45] / nev_sim$median_os[nev_sim$ALB == 35]
stopifnot(
  length(ratio_10) == 1L,
  abs(ratio_10 - exp(1.08)) < 1e-8,
  max(abs(nev_sim$sur - exp(-nev_sim$cumhazard))) < 1e-12
)
```

The albumin coefficient's units are self-checking. At a typical 38 g/L the model
gives a median OS of

```{r nonevaluable-units}
exp(0.681 + 0.108 * 38)
```

days -- a poor outcome, consistent with patients who never had an evaluable
postbaseline scan. Reading the same coefficient as "per g/dL" would give

```{r nonevaluable-units-wrong}
exp(0.681 + 0.108 * 3.8)
```

days, which is impossible. Supplementary Table S2 labels the row "Baseline
albumin (g/L)", and the arithmetic confirms it.

## An external check on the absolute scale (Supplementary Figure S3)

Supplementary Figure S3 is the paper's own visual predictive check for this
model: observed Kaplan-Meier curves for the nonevaluable patients in each arm,
overlaid with the model's prediction band. Locating the axis tick marks in the
figure and interpolating between them, the motesanib curve crosses S(t) = 0.5
at about 70 days and the placebo curve at about 125 days (the placebo step sits
flat on 0.5 from about 117 to 129 days, which sets the resolution of the
reading). Digitizing a printed step function is evidence about the figure, not
about the underlying data, so nothing below is asserted on those numbers
tightly -- the assertions are stated with enough margin to absorb a +/- 25%
reading error.

What *is* robust to that reading error is the scale. Serum albumin in
advanced-cancer patients runs roughly 30-45 g/L. Under the labelled g/L reading
that range brackets both digitized medians; under a g/dL misreading the model
misses by a factor of about fifty, which no plausible digitizing error can
close.

```{r nonevaluable-scale-check}
med_os_alb <- function(alb) exp(0.681 + 0.108 * alb)

scale_tbl <- tibble::tibble(
  Reading = c("g/L (as labelled)", "g/dL (the misreading)"),
  `Plausible albumin range` = c("30 to 45 g/L", "3.0 to 4.5 g/dL"),
  `Model median OS (days)` = c(
    paste(round(med_os_alb(c(30, 45))), collapse = " to "),
    paste(round(med_os_alb(c(3.0, 4.5)), 1), collapse = " to ")
  ),
  `At a typical 40 g/L (4.0 g/dL)` = round(med_os_alb(c(40, 4.0)), 1)
)
knitr::kable(scale_tbl)

stopifnot(
  # Across the plausible albumin range the g/L reading brackets both digitized
  # per-arm medians (about 70 days motesanib, 125 days placebo), with margin on
  # both sides rather than a bound that tracks the point estimate.
  med_os_alb(30) < 70, med_os_alb(45) > 125,
  # The g/dL reading is ~50-fold too low -- not a borderline call.
  med_os_alb(4.0) < 5
)
```

The albumin imbalance the paper invokes is also the right size: inverting the
regression at the two digitized medians implies

```{r nonevaluable-implied-alb}
implied_alb <- (log(c(motesanib = 70, placebo = 125)) - 0.681) / 0.108
round(implied_alb, 1)

stopifnot(
  # Both implied values land inside the physiologic albumin range, so the g/L
  # reading is self-consistent. A +/- 25% error on either digitized median
  # moves the implied albumin by only about 2.7 g/L, well inside this window.
  all(implied_alb > 25), all(implied_alb < 50),
  # The paper's own explanation: the placebo arm did better because its
  # nonevaluable patients started with HIGHER baseline albumin.
  implied_alb[["placebo"]] > implied_alb[["motesanib"]]
)
```

g/L, i.e. an arm difference of about
`r round(diff(unname(implied_alb)), 1)` g/L -- a modest, entirely plausible
baseline imbalance, which is precisely the explanation Claret 2014 offers for
the better outcome in the placebo arm among these patients. Under the g/dL
misreading the same two medians would demand baseline albumins of
`r round(implied_alb[["motesanib"]], 1)` and `r round(implied_alb[["placebo"]], 1)`
g/dL, roughly eight-fold above any physiologic value.

```{r fig-nonevaluable}
nev_curves <- tidyr::expand_grid(
  ALB = c(25, 30, 35, 40, 45),
  time = seq(1, 1200, by = 5)
) |>
  dplyr::mutate(
    sur = 1 - pnorm((log(time) - (0.681 + 0.108 * ALB)) / exp(0.502)),
    Albumin = factor(paste(ALB, "g/L"), levels = paste(c(25, 30, 35, 40, 45), "g/L"))
  )

ggplot(nev_curves, aes(time, sur, colour = Albumin)) +
  geom_line(linewidth = 0.7) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Time (days)", y = "Survival probability",
       title = "Nonevaluable patients: survival by baseline albumin",
       caption = paste(
         "Compare Supplementary Figure S3 of Claret 2014, the paper's visual",
         "predictive check for this model against the observed per-arm",
         "Kaplan-Meier curves.")) +
  theme_bw()
```

This reproduces the paper's observation that an imbalance in baseline albumin
could account for the better outcome seen in the CP-plus-placebo arm than in the
CP-plus-motesanib arm among nonevaluable patients.

# Why no NCA

Neither model contains a pharmacokinetic compartment. The simplified TGI model
is explicitly *not* exposure-driven -- motesanib's 125 mg once-daily regimen
enters only through the estimated `kdrug` and `lambda` -- and the OS models
consume covariates, not concentrations. There is therefore no concentration-time
profile to run through PKNCA. Validation here is by structural identity and by
comparison against the paper's published figures and tables, following the same
pattern as `vignette("Struemper_2025_tumorsize_OS_nsclc")` and
`vignette("Nagy_2017_obiltoxaximab")`.

# Assumptions and deviations

1. **Supplementary Table S1 transposes the KDE and lambda estimates; this
   extraction exchanges them.** Table S1 prints KDE = 0.0193 and lambda = 0.0103
   per day; the model file uses KDE = 0.0103 and lambda = 0.0193. **Both numbers
   are verbatim from Table S1 -- only their row assignment changes, and no value
   is tuned toward any target.** The full argument, with assertions, is in the
   section "Supplementary Table S1 transposes KDE and lambda" above. In brief:

   * The decisive test is eta-free. The TS model is a single pooled fit with no
     treatment-arm term, so its typical value must lie between the two published
     per-arm medians. As printed it gives a typical TTG of 281 days and a week-8
     TSratio of 0.467, outside the published 93-125 day and 0.69-0.77 intervals
     on both counts; corrected it gives 117 days and 0.746, inside both. Because
     the typical value contains no random effects, no eta-correlation or
     empirical-Bayes-shrinkage argument can account for the gap.
   * Forward Monte Carlo over the Table S1 IIV corroborates on the lower tail:
     the corrected reading gives a median TTG of `r round(s1_ttg_md)` days and a
     TSratio 2.5th percentile near 0.27 against a published 0.24-0.27, while the
     printed reading gives 257 days and 0.05.
   * Feeding the published TTG medians forward through the Table 3 regression
     reproduces Table 5's simulated survival (445 days computed vs 450 published
     on placebo); the printed Table S1 TTG implies roughly 865 days.

   A prior draft of this vignette attributed the same discrepancy to the
   difference between a forward marginal draw and a distribution of correlated
   empirical-Bayes estimates, and left Table S1 unchanged. The eta-free typical-
   value test above rules that explanation out.

   This has not been confirmed with the authors and no erratum to the paper is
   known. Users who need the printed Table S1 values can restore them with
   `rxode2::ini()`, at the cost of no longer matching the paper's Results.
2. **The TS sub-model has no treatment-arm term.** It is a single pooled fit
   across both MONET1 arms, so a forward simulation with freshly drawn etas
   produces the *same* TS and TTG distribution for motesanib and placebo. The
   paper's arm difference arises entirely from the individual empirical-Bayes
   estimates, and its trial simulations "were conditioned on 'predicted' tumor
   reductions and patient characteristics of MONET1 data". Consequently the
   packaged model cannot generate the published hazard ratios (Table 4: 0.88
   full population, 0.76 Asian) or the Table 5 power calculations de novo --
   those require the individual-level MONET1 data, which are not public. Neither
   is asserted in this vignette.
3. **IIV scale.** Supplementary Table S1's fourth column is headed "SD" and its
   abbreviation line defines it as the "standard deviation of inter-patient
   variability in diagonal covariance matrix". The values are therefore squared
   to give the variances nlmixr2's `eta ~ variance` syntax expects. This choice
   does not affect any median, so the checks in this vignette cannot discriminate
   it; the table's own wording is the evidence.
4. **Smoking-history polarity.** Table 3's coefficient (-0.264) is defined for
   "former and current smokers as compared with those who had never smoked",
   i.e. on an ever-smoker indicator. The canonical register column is
   `SMOKE_NEVER`, so `model()` forms the ever-smoker indicator as
   `1 - SMOKE_NEVER`. This keeps both the published intercept (1.079) and the
   published coefficient intact, preserving their correspondence with the
   Supplementary Table S3 covariance matrix; re-parameterising directly onto
   `SMOKE_NEVER` would have required shifting the intercept to 0.815.
5. **Baseline TS in the OS model.** The paper's OS regression uses observed
   baseline tumor size. The joint model uses the TS sub-model's own individual
   `rbase`, following the convention of
   `modellib("Struemper_2025_tumorsize_OS_nsclc")`. The two coincide at the
   typical value (90 mm) and differ only by estimation shrinkage.
6. **Nonevaluable cohort size.** Supplementary Table S2 is titled "... for
   Non-evaluable Patients (n = 146)", but the main text reports 156 nonevaluable
   patients (88 motesanib + 68 placebo, and separately 115 + 41). The paper does
   not reconcile the two; the most likely explanation is that 10 patients lacked
   a baseline albumin measurement. `population$n_subjects` is set to 146 (the
   number the regression was actually fitted on); the 14% correction applied in
   the simulations uses 156.
7. **Supplementary Figure S1 axis labels.** The goodness-of-fit panels label
   both axes "(cm)", but the Methods, Table S1 and the plotted range (0-420 with
   a typical baseline of 90) are all mm. Treated as a figure-labelling error;
   the models use mm throughout.
8. **`log(TTG + 100)` is left unguarded.** The published model is undefined for
   TTG <= -100 days and the paper specifies no fallback, so none is invented.
   The probability is quantified above (about
   `r sprintf("%.1f%%", 100 * p_undefined)` under the published IIV) rather than
   silently clipped.
9. **No IIV or residual error on either OS model.** Neither Table 3 nor
   Supplementary Table S2 reports any; the AFT scale parameter is the sole
   source of between-subject survival variability. No etas are invented.
10. **The historical OS model is not extracted.** Claret 2014 evaluates the
    earlier first-line NSCLC OS model of Wang et al. (its reference 7) but does
    not reprint its parameter estimates, so it cannot be sourced from this paper.
11. **TTG time unit: the Methods say weeks, everything else says days.** The
    Methods define metric 2 as "Time to tumor growth (TTG), the time to nadir
    TS (in weeks)". Four other locations contradict that word: the Methods
    themselves state that "the parameters KL, KDE, and `lambda` have inverse of
    time units (e.g., day^-1)", so `(log(KDE) - log(KL)) / lambda` is in days;
    Table 3 heads the row `Log(TTG*(day))`; Table 3 footnote c states that
    survival time was analysed in days; and the Results report the TTG medians
    as "125 (28-399) vs. 93 (-6 to 312) days". TTG is therefore taken as
    **days** throughout, as is the shift constant in `log(TTG + 100)`. Reading
    "(in weeks)" literally would rescale TTG by a factor of 7 and, through the
    0.97695 coefficient on `log(TTG*)`, would move predicted median survival by
    a factor of roughly 6, so the distinction is load-bearing rather than
    cosmetic.
12. **Figure-digitized values are used for narrative only, never for
    parameters.** Two numbers in this vignette were read off a printed figure
    rather than taken from the text or tables: the Supplementary Figure S3
    per-arm median survival times for nonevaluable patients (about 70 days on
    motesanib and about 125 days on placebo). Both were obtained by locating
    the printed axis tick marks in the page image and interpolating linearly
    between them, not by eye. Figure 1's TTG-quartile medians are deliberately
    *not* quoted numerically -- the second- and third-quartile curves cross
    near S(t) = 0.5, so those two crossing times cannot be separated reliably
    from the printed figure, and only the monotone ordering is claimed. Every
    `ini()` value in both model files comes from Table 3, Supplementary Table
    S1 or Supplementary Table S2; no parameter is derived from a figure. The
    assertions that reference the two digitized readings are stated as windows
    wide enough to absorb a +/- 25% reading error.
