---
title: "Week-8 disease control rate as a predictor of OS and PFS in NSCLC (Claret 2014)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Week-8 disease control rate as a predictor of OS and PFS in NSCLC (Claret 2014)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
set.seed(20140301)
```

# Model and source

```{r model-load, include = FALSE}
mod_os  <- rxode2::rxode(readModelDb("Claret_2014_DCR_OS_nsclc"))
mod_pfs <- rxode2::rxode(readModelDb("Claret_2014_DCR_PFS_nsclc"))
```

* Citation: `r mod_os$reference`
* Article: [J Clin Pharmacol. 2014;54(3):253-257](https://doi.org/10.1002/jcph.191)

This paper contributes **two** models to `nlmixr2lib`, matching the two
independent `survreg` regressions the authors report:

* `modellib("Claret_2014_DCR_OS_nsclc")` -- overall survival (paper Table 2).
* `modellib("Claret_2014_DCR_PFS_nsclc")` -- progression-free survival (paper Table 3).

Both are accelerated failure time (AFT) log-normal survival regressions with no
PK input, no ODE state, no inter-individual variability and no residual error.
They are algebraic: given a patient's covariates they return the location
parameter of `log(T)`, the median survival time, and the survivor function
directly.

* OS description: `r mod_os$description`
* PFS description: `r mod_pfs$description`

# Population

The models were fit to the **evaluable** population of two bevacizumab studies
in first-line advanced non-small cell lung cancer (NSCLC): the Phase III E4599
study in 878 Western patients (bevacizumab + carboplatin/paclitaxel versus
carboplatin/paclitaxel alone) and the single-arm Phase IV SAiL study in 198
Chinese patients. To be evaluable a patient needed at least one response
assessment between weeks 3 and 14, and the assessment closest to week 8 was the
one used; 629 of 878 Western patients (72%) and 145 of 198 Chinese patients
(73%) qualified, giving **n = 774**.

Patient characteristics were limited to two well-established prognostic factors:
baseline serum albumin (< 3.5 g/dL versus >= 3.5 g/dL) and ECOG performance
status (0 versus >= 1). Low albumin was much more common in the Western study
(29%) than the Chinese one (6%), while ECOG >= 1 was slightly more common in the
Chinese study (78% versus 60%).

Week-8 response categories in the evaluable population reproduce the paper's
Table 1:

```{r table1}
table1 <- tibble::tribble(
  ~Study,    ~Arm,             ~CR, ~PR,  ~SD,  ~PD, ~Total,
  "Western", "Chemo",           2L,  67L, 152L,  99L,   320L,
  "Western", "Bev + Chemo",     4L, 133L, 128L,  44L,   309L,
  "Chinese", "Bev + Chemo",     2L,  67L,  69L,   7L,   145L
) |>
  dplyr::mutate(DCR = (CR + PR + SD) / Total)

stopifnot(with(table1, all(CR + PR + SD + PD == Total)))

table1 |>
  dplyr::mutate(DCR = sprintf("%.1f%%", 100 * DCR)) |>
  knitr::kable(caption = "Week-8 response categories in the evaluable population (reproduces Claret 2014 Table 1); DCR = (CR + PR + SD) / Total.")
```

The paper's observation that disease control was higher in the Chinese study
(95%) than among Western bevacizumab-treated patients (85%) is recovered
exactly by the last column.

The same information is available programmatically from each model's
`population` metadata (`mod_os$population`).

# Source trace

Every `ini()` value is carried with an in-file comment in
`inst/modeldb/therapeuticArea/oncology/Claret_2014_DCR_OS_nsclc.R` and
`..._PFS_nsclc.R`. They are collected here for review.

| Model | Parameter | Value | SE | Source location |
|---|---|---|---|---|
| OS | `mu_os_int` | 1.855 | 0.0613 | Table 2, row "Intercept" |
| OS | `e_dcr_mu_os` | 0.939 | 0.0646 | Table 2, row "DCR" |
| OS | `e_alb_lt35_mu_os` | -0.270 | 0.0601 | Table 2, row "Albumin < 3.5 (g/dL)" |
| OS | `lsigma_os` | -0.370 | 0.0308 | Table 2, row "Log (scale)" |
| PFS | `mu_pfs_int` | 0.466 | 0.0477 | Table 3, row "Intercept" |
| PFS | `e_dcr_mu_pfs` | 1.378 | 0.0525 | Table 3, row "DCR" |
| PFS | `e_bev_mu_pfs` | 0.232 | 0.0429 | Table 3, row "Bevacizumab" |
| PFS | `lsigma_pfs` | -0.641 | 0.0282 | Table 3, row "Log (scale)" |
| both | AFT log-normal form, `survreg` | n/a | | Methods, "Survival Model Development"; Results, "Overall Survival Model" and "PFS Model" |
| both | Week-8 DCR definition (CR + PR + SD vs PD) | n/a | | Methods, "Response Metrics" |
| both | Albumin dichotomy at 3.5 g/dL; ECOG dichotomy at 1 | n/a | | Methods, "Patient Characteristics" |
| both | Censoring: study duration ~ Uniform(8, 30) months | n/a | | Methods, "Survival Model Development" |

## Sign recovery for the negative coefficients

Three of the eight estimates are negative. Their minus signs are typeset as
U+2212 and are **dropped by both** the trimmed-markdown conversion of the PDF
and by `pdftotext -layout`, so they cannot be taken from a text extraction. They
were recovered from the tables' own Wald statistics -- `z` equals estimate
divided by SE on every row, so the sign of `z` supplies the sign of the estimate
-- and then confirmed by reading the rendered page image.

```{r sign-recovery}
sign_check <- tibble::tribble(
  ~Model, ~Parameter,           ~Estimate, ~SE,     ~z_published,
  "OS",   "Intercept",            1.855,   0.0613,   30.2,
  "OS",   "DCR",                  0.939,   0.0646,   14.5,
  "OS",   "Albumin < 3.5 g/dL",  -0.270,   0.0601,   -4.5,
  "OS",   "Log (scale)",         -0.370,   0.0308,  -12.0,
  "PFS",  "Intercept",            0.466,   0.0477,    9.8,
  "PFS",  "DCR",                  1.378,   0.0525,   26.2,
  "PFS",  "Bevacizumab",          0.232,   0.0429,    5.4,
  "PFS",  "Log (scale)",         -0.641,   0.0282,  -22.7
) |>
  dplyr::mutate(z_recomputed = Estimate / SE,
                rel_diff = abs(z_recomputed / z_published - 1),
                agrees = sign(z_recomputed) == sign(z_published) & rel_diff < 0.01)

# Every row must reproduce the published Wald statistic, sign included.
#
# The test is stated relatively rather than absolutely because the published
# estimate, SE and z are each rounded: the largest honest disagreement across
# the eight rows is 0.31% (the intercept, 30.261 recomputed vs 30.2 printed),
# so a 1% tolerance absorbs rounding. It remains an extremely strict test of
# the signs -- a flipped sign changes z by twice its magnitude, i.e. a relative
# difference of 200%, which is 640-fold above the threshold and could not be
# mistaken for rounding.
stopifnot(all(sign_check$agrees), max(sign_check$rel_diff) < 0.01)

sign_check |>
  dplyr::mutate(z_recomputed = round(z_recomputed, 2),
                rel_diff = sprintf("%.2f%%", 100 * rel_diff)) |>
  dplyr::rename("Published z" = z_published, "Estimate / SE" = z_recomputed,
                "Relative difference" = rel_diff, "Signs agree" = agrees) |>
  knitr::kable(caption = "Sign recovery: estimate / SE reproduces the published Wald statistic for all eight parameters, which fixes the sign of each.")
```

The negative albumin coefficient is independently corroborated by the paper's
prose -- "the probability to survive decreases in patients with low baseline
albumin" -- and the positive bevacizumab coefficient by "the probability of
progression or death decreases in patients with disease control and with
bevacizumab treatment".

# Mechanism in one paragraph

`survreg`'s log-normal fit is the AFT form `log(T) = mu + sigma * eps` with
`eps ~ N(0, 1)`, so `T` is log-normal with median `exp(mu)` and the table's
`Log (scale)` row is `log(sigma)`. Each model makes `mu` linear in its retained
binary covariates and leaves `sigma` covariate-free, which means every covariate
acts as a pure **multiplier on median survival**: disease control at week 8
multiplies median OS by `exp(0.939) = 2.56` and median PFS by
`exp(1.378) = 3.97`; low albumin multiplies median OS by `exp(-0.270) = 0.76`;
bevacizumab multiplies median PFS by `exp(0.232) = 1.26`. The survivor function
is `S(t) = 1 - Phi((log(t) - mu) / sigma)`.

The structural difference between the two models is the paper's main result. In
the OS model, bevacizumab treatment was significant univariately but dropped out
of the multivariate model because week-8 disease control fully captured it, so
the final OS model carries **no drug term at all** and the authors describe it
as drug independent. In the PFS model, disease control did *not* fully capture
the bevacizumab effect, so a separate `DRUG_BEV` term is retained. Chinese
ethnicity was significant univariately for both endpoints and eliminated from
both final models, with no DCR-by-ethnicity interaction -- the paper's title
result.

# Structural identities

Before comparing against anything published, check that the packaged models
compute what the AFT algebra says they should. These are exact identities, so
they are asserted tightly.

```{r identities}
# The models are algebraic: a plain data frame of id / time / covariates is a
# valid event table. There are no ODE states, so there is no `cmt` and no
# `evid`, and no `omega = NA` (these models declare no etas at all).
grid_os <- tidyr::expand_grid(RESP_DCR = c(0, 1), ALB_LT35 = c(0, 1)) |>
  dplyr::mutate(id = dplyr::row_number()) |>
  tidyr::expand_grid(time = c(1, 3, 6, 12, 24, 36))

sim_os <- as.data.frame(rxode2::rxSolve(mod_os, events = grid_os,
                                        keep = c("RESP_DCR", "ALB_LT35")))

# 1. median_os is exactly exp(mu_os).
stopifnot(max(abs(sim_os$median_os - exp(sim_os$mu_os))) < 1e-10)

# 2. cumhazard_os is exactly -log(sur_os).
stopifnot(max(abs(sim_os$cumhazard_os + log(sim_os$sur_os))) < 1e-8)

# 3. sur_os matches the closed-form log-normal survivor function.
sur_closed <- 1 - pnorm((log(sim_os$time) - sim_os$mu_os) / sim_os$sigma_os)
stopifnot(max(abs(sim_os$sur_os - sur_closed)) < 1e-6)

# 4. Covariate effects are exact multipliers on the median. Compare the four
#    covariate cells by key, never by row position.
med <- sim_os |>
  dplyr::distinct(RESP_DCR, ALB_LT35, median_os)
cell <- function(d, a) {
  v <- med$median_os[med$RESP_DCR == d & med$ALB_LT35 == a]
  if (length(v) != 1L) stop("no unique cell for DCR=", d, " ALB=", a)
  v
}
stopifnot(
  abs(cell(1, 0) / cell(0, 0) - exp(0.939))  < 1e-8,   # DCR effect
  abs(cell(1, 1) / cell(1, 0) - exp(-0.270)) < 1e-8,   # albumin effect
  # The model is additive on the log scale, so the two effects must be
  # exactly multiplicative on the median with no interaction term.
  abs(cell(1, 1) / cell(0, 0) - exp(0.939 - 0.270)) < 1e-8
)

# 5. The survivor function at t = median must be exactly 0.5 -- a definitional
#    check that exercises the pnorm path against the mu/sigma path.
at_median <- med |>
  dplyr::mutate(id = dplyr::row_number(), time = median_os) |>
  dplyr::select(id, time, RESP_DCR, ALB_LT35)
sur_at_median <- as.data.frame(rxode2::rxSolve(mod_os, events = at_median))$sur_os
stopifnot(max(abs(sur_at_median - 0.5)) < 1e-6)

med |>
  dplyr::arrange(RESP_DCR, ALB_LT35) |>
  dplyr::mutate(median_os = round(median_os, 2)) |>
  dplyr::rename("Week-8 disease control" = RESP_DCR,
                "Albumin < 3.5 g/dL" = ALB_LT35,
                "Median OS (months)" = median_os) |>
  knitr::kable(caption = "Median overall survival by covariate cell. The 2.56-fold disease-control effect and 0.76-fold low-albumin effect are exact multipliers.")
```

The same identities hold for the PFS model.

```{r identities-pfs}
grid_pfs <- tidyr::expand_grid(RESP_DCR = c(0, 1), DRUG_BEV = c(0, 1)) |>
  dplyr::mutate(id = dplyr::row_number()) |>
  tidyr::expand_grid(time = c(1, 3, 6, 12, 24))

sim_pfs <- as.data.frame(rxode2::rxSolve(mod_pfs, events = grid_pfs,
                                         keep = c("RESP_DCR", "DRUG_BEV")))

stopifnot(
  max(abs(sim_pfs$median_pfs - exp(sim_pfs$mu_pfs))) < 1e-10,
  max(abs(sim_pfs$cumhazard_pfs + log(sim_pfs$sur_pfs))) < 1e-8,
  max(abs(sim_pfs$sur_pfs -
          (1 - pnorm((log(sim_pfs$time) - sim_pfs$mu_pfs) / sim_pfs$sigma_pfs)))) < 1e-6
)

med_pfs <- sim_pfs |> dplyr::distinct(RESP_DCR, DRUG_BEV, median_pfs)
cell_pfs <- function(d, b) {
  v <- med_pfs$median_pfs[med_pfs$RESP_DCR == d & med_pfs$DRUG_BEV == b]
  if (length(v) != 1L) stop("no unique cell for DCR=", d, " BEV=", b)
  v
}
stopifnot(
  abs(cell_pfs(1, 0) / cell_pfs(0, 0) - exp(1.378)) < 1e-8,
  abs(cell_pfs(0, 1) / cell_pfs(0, 0) - exp(0.232)) < 1e-8
)

med_pfs |>
  dplyr::arrange(RESP_DCR, DRUG_BEV) |>
  dplyr::mutate(median_pfs = round(median_pfs, 2)) |>
  dplyr::rename("Week-8 disease control" = RESP_DCR,
                "Bevacizumab" = DRUG_BEV,
                "Median PFS (months)" = median_pfs) |>
  knitr::kable(caption = "Median progression-free survival by covariate cell. Disease control is a 3.97-fold multiplier; bevacizumab adds a further 1.26-fold on top of it.")
```

# Replicate Figures 1 and 3: survival by week-8 response category

Figures 1 and 3 of Claret 2014 show Kaplan-Meier curves stratified by week-8
response category with model-predicted 95% prediction intervals overlaid. The
packaged models produce the underlying predicted survivor curves directly. The
paper's models use the dichotomised DCR, so patients with CR, PR or SD share one
predicted curve and patients with PD share the other; the low-albumin split is
shown for OS because it is the model's other retained covariate.

```{r fig-1-3, fig.cap = "Model-predicted survivor functions by week-8 disease control. Left: overall survival, split by baseline albumin (the OS model's second covariate). Right: progression-free survival, split by bevacizumab treatment (the PFS model's second covariate). Replicates the model-predicted bands of Figure 1 and Figure 3 of Claret 2014."}
tgrid <- seq(0.25, 36, by = 0.25)

curve_os <- tidyr::expand_grid(RESP_DCR = c(0, 1), ALB_LT35 = c(0, 1)) |>
  dplyr::mutate(id = dplyr::row_number()) |>
  tidyr::expand_grid(time = tgrid)
curve_os <- as.data.frame(
  rxode2::rxSolve(mod_os, events = curve_os, keep = c("RESP_DCR", "ALB_LT35"))
) |>
  dplyr::transmute(
    Endpoint = "Overall survival",
    time, sur = sur_os,
    stratum = paste0(ifelse(RESP_DCR == 1, "Disease control", "Progressive disease"),
                     ifelse(ALB_LT35 == 1, ", albumin < 3.5", ", albumin >= 3.5"))
  )

curve_pfs <- tidyr::expand_grid(RESP_DCR = c(0, 1), DRUG_BEV = c(0, 1)) |>
  dplyr::mutate(id = dplyr::row_number()) |>
  tidyr::expand_grid(time = tgrid)
curve_pfs <- as.data.frame(
  rxode2::rxSolve(mod_pfs, events = curve_pfs, keep = c("RESP_DCR", "DRUG_BEV"))
) |>
  dplyr::transmute(
    Endpoint = "Progression-free survival",
    time, sur = sur_pfs,
    stratum = paste0(ifelse(RESP_DCR == 1, "Disease control", "Progressive disease"),
                     ifelse(DRUG_BEV == 1, ", bevacizumab", ", chemo alone"))
  )

dplyr::bind_rows(curve_os, curve_pfs) |>
  ggplot(aes(time, sur, colour = stratum)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~Endpoint) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Time (months)", y = "Predicted survival probability", colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom", legend.direction = "vertical")
```

The separation between the disease-control and progressive-disease curves is
much wider for PFS than for OS, which is the quantitative form of the paper's
finding that week-8 response is a more direct read-out of progression than of
death (the DCR coefficient is 1.378 for PFS versus 0.939 for OS).

# Replicate Figure 2: the bevacizumab hazard ratio

Figure 2 of Claret 2014 is the paper's own posterior predictive check and is the
best available answer key for these models: the authors simulated the Western
E4599 study 1,000 times, sampling model parameters from their estimates and
standard errors on each replicate, applied the published censoring scheme, and
recorded the distribution of the bevacizumab hazard ratio.

The replication below follows the paper's described procedure. Bevacizumab is
absent from the OS model, so the *only* route by which it can produce an OS
hazard ratio is by raising the probability of week-8 disease control -- which is
precisely the paper's claim that DCR fully captures the bevacizumab OS effect.
For PFS, bevacizumab acts through both that route and the retained `DRUG_BEV`
term.

```{r hr-replication}
n_per_arm <- 200L        # capped per the 200-participants-per-arm vignette policy
n_rep     <- 1000L       # the paper simulated 1,000 replicates

# Arm-level covariate distributions, all from the paper.
dcr_chemo <- with(table1, (CR + PR + SD)[Arm == "Chemo"] / Total[Arm == "Chemo"])
dcr_bev   <- with(table1, (CR + PR + SD)[Study == "Western" & Arm == "Bev + Chemo"] /
                          Total[Study == "Western" & Arm == "Bev + Chemo"])
p_alb_low <- 0.29        # Methods: 29% of Western patients had albumin < 3.5 g/dL
stopifnot(abs(dcr_chemo - 0.691) < 0.001, abs(dcr_bev - 0.858) < 0.001)

# Published Table 2 / Table 3 estimates and standard errors, for the
# parameter-uncertainty draw the paper describes.
est_os  <- c(int =  1.855, dcr = 0.939, alb = -0.270, lsc = -0.370)
se_os   <- c(int = 0.0613, dcr = 0.0646, alb = 0.0601, lsc = 0.0308)
est_pfs <- c(int =  0.466, dcr = 1.378, bev =  0.232, lsc = -0.641)
se_pfs  <- c(int = 0.0477, dcr = 0.0525, bev = 0.0429, lsc = 0.0282)

one_replicate <- function(...) {
  o <- rnorm(4, est_os,  se_os)
  p <- rnorm(4, est_pfs, se_pfs)
  arm <- rep(c(0L, 1L), each = n_per_arm)
  # Bevacizumab raises the probability of week-8 disease control.
  dcr <- rbinom(2L * n_per_arm, 1L, ifelse(arm == 1L, dcr_bev, dcr_chemo))
  alb <- rbinom(2L * n_per_arm, 1L, p_alb_low)

  t_os  <- rlnorm(2L * n_per_arm, o[1] + o[2] * dcr + o[3] * alb, exp(o[4]))
  t_pfs <- rlnorm(2L * n_per_arm, p[1] + p[2] * dcr + p[3] * arm, exp(p[4]))

  # Methods: censoring simulated from Uniform(8, 30) months.
  cens <- runif(2L * n_per_arm, 8, 30)
  cox <- function(tt) {
    ev <- as.integer(tt <= cens)
    unname(exp(coef(survival::coxph(survival::Surv(pmin(tt, cens), ev) ~ arm))))
  }
  c(os = cox(t_os), pfs = cox(t_pfs))
}

hr <- as.data.frame(t(vapply(seq_len(n_rep), one_replicate, numeric(2))))
```

```{r fig-2, fig.cap = "Distribution of the model-predicted bevacizumab hazard ratio over 1,000 simulated replicates of the Western E4599 study. Solid vertical line: observed HR. Dashed lines: 2.5th and 97.5th percentiles of the predicted distribution. Replicates Figure 2 of Claret 2014 (OS panel) and its PFS analogue."}
observed <- tibble::tibble(Endpoint = c("Overall survival", "Progression-free survival"),
                           obs = c(0.77, 0.58))

hr_long <- hr |>
  tidyr::pivot_longer(everything(), names_to = "key", values_to = "HR") |>
  dplyr::mutate(Endpoint = ifelse(key == "os", "Overall survival",
                                  "Progression-free survival"))

pi_lines <- hr_long |>
  dplyr::group_by(Endpoint) |>
  dplyr::summarise(lo = quantile(HR, 0.025), hi = quantile(HR, 0.975), .groups = "drop") |>
  tidyr::pivot_longer(c(lo, hi), values_to = "x", names_to = "bound")

ggplot(hr_long, aes(HR)) +
  geom_histogram(bins = 45, fill = "grey70", colour = "white") +
  geom_vline(data = pi_lines, aes(xintercept = x), linetype = "dashed") +
  geom_vline(data = observed, aes(xintercept = obs), colour = "firebrick",
             linewidth = 0.9) +
  facet_wrap(~Endpoint, scales = "free") +
  labs(x = "Bevacizumab hazard ratio (bevacizumab + chemo vs chemo)",
       y = "Replicates") +
  theme_bw()
```

## Comparison against the published predictive check

```{r hr-compare}
hr_summary <- hr_long |>
  dplyr::group_by(Endpoint) |>
  dplyr::summarise(
    sim_median = median(HR),
    sim_lo     = quantile(HR, 0.025),
    sim_hi     = quantile(HR, 0.975),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    tibble::tibble(
      Endpoint  = c("Overall survival", "Progression-free survival"),
      pub_pred  = c(0.84, 0.59),
      pub_lo    = c(0.71, 0.49),
      pub_hi    = c(0.98, 0.72),
      pub_obs   = c(0.77, 0.58)
    ),
    by = "Endpoint"
  ) |>
  dplyr::mutate(pct_diff = 100 * (sim_median - pub_pred) / pub_pred)

# Structural assertion on the CENTRE of the predicted distribution. A wrong
# sign, a mis-transcribed coefficient or a scale-vs-log(scale) confusion moves
# the median hazard ratio far outside these bounds.
#
# The bound is 5%, against achieved differences of -2.5% (OS) and 0.0% (PFS).
# All randomness here comes from base R's RNG under a fixed seed (the models
# themselves are deterministic -- they declare no etas), so this is reproducible
# across rxode2 versions, unlike an assertion on an rxode2-drawn cohort. The
# Monte Carlo standard error on the median of 1,000 replicates is about 0.5%,
# so 5% leaves roughly five standard errors of headroom over the residual.
stopifnot(all(abs(hr_summary$pct_diff) < 5))

# The paper's OBSERVED hazard ratios must fall inside the simulated prediction
# interval -- that is exactly the claim Figure 2 makes.
stopifnot(with(hr_summary, all(pub_obs > sim_lo & pub_obs < sim_hi)))

hr_summary |>
  dplyr::transmute(
    Endpoint,
    `Simulated HR (median)`      = sprintf("%.2f", sim_median),
    `Simulated 95% PI`           = sprintf("%.2f-%.2f", sim_lo, sim_hi),
    `Published predicted HR`     = sprintf("%.2f", pub_pred),
    `Published 95% PI`           = sprintf("%.2f-%.2f", pub_lo, pub_hi),
    `Observed HR`                = sprintf("%.2f", pub_obs),
    `Difference vs published (%)` = sprintf("%+.1f", pct_diff)
  ) |>
  knitr::kable(caption = "Simulated versus published bevacizumab hazard ratios in the Western evaluable population. Published values are from the Claret 2014 Abstract and Results.")
```

The simulated median hazard ratios reproduce the published predictions to within
a few percent for both endpoints, and the observed hazard ratios fall inside the
simulated prediction intervals. Because the OS model contains no bevacizumab
term, its hazard ratio of roughly 0.82 is generated **entirely** by the higher
week-8 disease-control rate in the bevacizumab arm -- a direct demonstration of
the paper's central claim.

The simulated prediction intervals are somewhat wider than the published ones.
That is expected and is not a model discrepancy: this vignette caps the cohort
at 200 patients per arm under the library's simulation-size policy, whereas the
paper simulated the full evaluable Western cohort of 320 + 309 = 629 patients.
The interval width scales with cohort size; the median, which is what the
assertions test, does not.

# Arm-level median survival

A secondary, qualitative comparison against the medians reported for the two
underlying trials.

```{r arm-medians}
arms <- tibble::tibble(
  Arm       = c("E4599 chemo", "E4599 bev + chemo", "SAiL bev + chemo"),
  dcr       = c(dcr_chemo, dcr_bev,
                with(table1, (CR + PR + SD)[Study == "Chinese"] / Total[Study == "Chinese"])),
  p_alb_low = c(0.29, 0.29, 0.06),     # Methods: 29% Western, 6% Chinese
  bev       = c(0, 1, 1)
)

# Median of the covariate mixture: solve S(t) = 0.5 over the 2x2 covariate cells
# weighted by their prevalence in the arm.
mixture_median <- function(mus, sigma, w) {
  stopifnot(abs(sum(w) - 1) < 1e-12)
  uniroot(function(tt) sum(w * (1 - pnorm((log(tt) - mus) / sigma))) - 0.5,
          c(0.01, 500))$root
}

sigma_os  <- exp(-0.370)
sigma_pfs <- exp(-0.641)

arm_medians <- arms |>
  dplyr::rowwise() |>
  dplyr::mutate(
    # cells, in order: (DCR=1,alb=0), (DCR=1,alb=1), (DCR=0,alb=0), (DCR=0,alb=1)
    w = list(c(dcr * (1 - p_alb_low), dcr * p_alb_low,
               (1 - dcr) * (1 - p_alb_low), (1 - dcr) * p_alb_low)),
    median_os  = mixture_median(1.855 + 0.939 * c(1, 1, 0, 0) - 0.270 * c(0, 1, 0, 1),
                                sigma_os, w),
    median_pfs = mixture_median(0.466 + 1.378 * c(1, 1, 0, 0) + 0.232 * bev,
                                sigma_pfs, w)
  ) |>
  dplyr::ungroup() |>
  dplyr::select(-w)

published_arm <- tibble::tibble(
  Arm        = c("E4599 chemo", "E4599 bev + chemo", "SAiL bev + chemo"),
  pub_os     = c(10.3, 12.3, 18.9),
  pub_pfs    = c(4.5, 6.2, 8.3)
)

# Directional assertions only -- see the caveat below on population mismatch.
# The model must order the arms the way the trials did.
chk <- dplyr::left_join(arm_medians, published_arm, by = "Arm")
stopifnot(
  # Bevacizumab arm outlives the chemotherapy arm on both endpoints.
  chk$median_os[chk$Arm == "E4599 bev + chemo"]  > chk$median_os[chk$Arm == "E4599 chemo"],
  chk$median_pfs[chk$Arm == "E4599 bev + chemo"] > chk$median_pfs[chk$Arm == "E4599 chemo"],
  # The Chinese arm outlives the Western bevacizumab arm, which the paper
  # attributes entirely to its higher DCR and lower rate of low albumin.
  chk$median_os[chk$Arm == "SAiL bev + chemo"]  > chk$median_os[chk$Arm == "E4599 bev + chemo"],
  chk$median_pfs[chk$Arm == "SAiL bev + chemo"] > chk$median_pfs[chk$Arm == "E4599 bev + chemo"]
)

chk |>
  dplyr::transmute(
    Arm,
    `Model median OS (mo)`  = sprintf("%.1f", median_os),
    `Trial median OS (mo)`  = sprintf("%.1f", pub_os),
    `Model median PFS (mo)` = sprintf("%.1f", median_pfs),
    `Trial median PFS (mo)` = sprintf("%.1f", pub_pfs)
  ) |>
  knitr::kable(caption = "Model-predicted arm medians versus the medians reported for E4599 and SAiL. Trial medians are for the ALL-RANDOMIZED populations; the model was fit to the week-8-evaluable subset, so exact agreement is not expected (see Assumptions).")
```

The model reproduces the ordering of all three arms and the direction of both
covariate effects. It over-predicts the two E4599 medians by roughly 1 month
each and under-predicts the SAiL median OS. Both residuals have the same
explanation and neither indicates a transcription error -- see below.

# Assumptions and deviations

* **Trial medians are not a like-for-like comparison.** The E4599 and SAiL
  medians quoted in the paper's introduction (OS 12.3 / 10.3 / 18.9 months; PFS
  6.2 / 4.5 months; SAiL time to progressive disease 8.3 months) are for the
  **all-randomized** populations, whereas both models were fit to the week-8
  **evaluable** subset (72% and 73% of enrolment). Patients who died or
  progressed before their week-8 assessment are excluded from the model's
  population by construction, which biases the model's arm medians upward
  relative to the trial medians -- exactly the direction seen for both E4599
  arms. The SAiL residual runs the other way; SAiL was a single-arm Phase IV
  study in a different care setting, and the paper itself only claims the
  covariates "explained" the longer Chinese survival, not that the magnitude was
  matched. The assertions in this section are therefore directional
  (arm ordering), not numeric. The numeric assertions in this vignette are all
  attached to the paper's own predictive check, which is computed on the
  evaluable population and is a like-for-like comparison.

* **Covariates are assumed independent within an arm.** The paper reports the
  week-8 DCR by arm (Table 1) and the low-albumin prevalence by study (Methods)
  but never cross-tabulates them, so the simulations here draw `RESP_DCR` and
  `ALB_LT35` independently. In reality low albumin is a poor-prognosis marker
  and is likely negatively associated with disease control, which would make the
  true joint distribution slightly more concentrated than the one simulated. The
  effect on the bevacizumab hazard ratio is second-order because albumin is
  balanced across the randomized arms.

* **Low-albumin prevalence is taken as arm-independent within E4599.** The paper
  reports 29% for the Western study as a whole; because E4599 was randomized,
  applying the same 29% to both arms is the appropriate assumption.

* **ECOG performance status is not simulated.** It was screened in both models
  and retained in neither, and no point estimate is published for its effect, so
  it cannot enter either model. It is recorded in each model file's
  `covariatesDataExcluded` list, together with bevacizumab treatment (OS model)
  and baseline albumin (PFS model).

* **Cohort size capped at 200 per arm.** The library policy caps simulated
  cohorts at 200 participants per arm; the paper simulated the full 320 + 309
  evaluable Western cohort. This widens the simulated prediction intervals
  relative to the published ones without shifting their centre, which is why the
  numeric assertions test the median hazard ratio rather than the interval
  bounds.

* **Sign recovery for three coefficients.** The minus signs on the OS albumin
  coefficient, the OS `Log (scale)` and the PFS `Log (scale)` are not recoverable
  from a text extraction of the PDF (both the trimmed markdown and
  `pdftotext -layout` drop the U+2212 glyph). They were recovered from the
  published Wald statistics, verified against the rendered page image, and the
  check is reproduced in the "Sign recovery" section above. All eight parameters
  satisfy `z = estimate / SE`.

* **No IIV and no residual error.** Neither table reports variance components.
  This is expected for a population-level parametric time-to-event regression:
  all subject-level variability is carried by the AFT scale `sigma`, so the
  models declare no `eta` terms. Consequently `rxode2::zeroRe()` and
  `omega = NA` are neither needed nor valid for these models.

* **Three new canonical covariate columns were registered** with this extraction,
  each a member of an established family in
  `inst/references/covariate-columns.md`: `RESP_DCR` (fourth member of the
  `RESP_<contrast>` RECIST-response family, alongside `RESP_SD`,
  `RESP_NONPDCR` and `RESP_RESPONDER`), `ALB_LT35` (threshold binarisation of
  the continuous `ALB` canonical, following the `AGE_GE70` / `MET_GE3` /
  `NTARGET_GE3` pattern) and `DRUG_BEV` (member of the `DRUG_<INN>` family
  alongside `DRUG_ORMU` and `DRUG_PRED`; unlike those head-to-head comparators
  its reference arm receives active chemotherapy, so it is not interchangeable
  with `ON_TREATMENT` or `PLACEBO`).

* **No PKNCA validation.** These models have no PK component, no dosing and no
  concentration-time curve, so non-compartmental analysis does not apply. The
  validation strategy is instead the exact structural identities of the AFT
  algebra plus the paper's own posterior predictive check on the bevacizumab
  hazard ratio.

* **The PFS model is not drug independent.** The paper states plainly that
  because week-8 disease control did not fully capture the bevacizumab effect on
  PFS, "the PFS model cannot be assumed to be drug independent and it has less
  value than the OS model." Users applying `Claret_2014_DCR_PFS_nsclc` to a
  non-bevacizumab agent should read that caveat first; the OS model carries no
  such restriction.
