---
title: "TB combination therapy MTP-GPDI in vitro (Clewe 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TB combination therapy MTP-GPDI in vitro (Clewe 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Clewe_2018_TB_MTP_GPDI_invitro"))$reference`
* Description: `r rxode2::rxode(readModelDb("Clewe_2018_TB_MTP_GPDI_invitro"))$description`
* Article: <https://doi.org/10.1093/jac/dkx380>

# Population

The model was developed from in vitro time-kill experiments using the
Mycobacterium tuberculosis Beijing VN 2002-1585 (BE1585) genotype strain,
cultured in Middlebrook 7H9 broth + 10% OADC + 0.5% glycerol + 0.02% Tween 20
under shaking conditions (96 rpm, 37 C) at the Erasmus Medical Centre,
Rotterdam, NL. Static drug concentrations were used as input to the PD
modelling; the stability of rifampicin, isoniazid, and ethambutol over the
6-day experiment allowed activity assessment without replenishment.

The limit of quantification was 5 CFU; data below LOQ were handled in NONMEM
with the M3 method. Tested concentrations (Clewe 2018 Figure 1): rifampicin
0.002 - 8 mg/L; isoniazid 0.01 - 40 mg/L; ethambutol 0.0078 - 32 mg/L. Duo
and trio combinations sampled cross-products of selected concentrations,
including clinically relevant unbound Cmax values (RIF 2, INH 10, EMB 8
mg/L).

The `population` metadata field of the model carries the same information
programmatically:

```{r population}
rxode2::rxode(readModelDb("Clewe_2018_TB_MTP_GPDI_invitro"))$population
```

# Source trace

Every `ini()` entry carries an in-file comment pointing to the Clewe 2018 (or
upstream Clewe 2016 MTP) source location. The table below collects them for
reviewer use.

| Block | Parameter | Value | Units | Source |
|-------|-----------|-------|-------|--------|
| MTP growth | `lkg`       | log(0.796)     | 1/day  | Clewe 2018 Table 1 (RSE 5%) |
| MTP growth | `lkfs_lin`  | log(1.66e-3) FIX | 1/day^2 | Clewe 2018 Table 1 footnote b (fixed from Clewe 2016) |
| MTP growth | `lkfn`      | log(8.97e-7) FIX | 1/day | Clewe 2018 Table 1 footnote b |
| MTP growth | `lksn`      | log(0.186) FIX   | 1/day | Clewe 2018 Table 1 footnote b |
| MTP growth | `lksf`      | log(0.0145) FIX  | 1/day | Clewe 2018 Table 1 footnote b |
| MTP growth | `lkns`      | log(1.23e-3) FIX | 1/day | Clewe 2018 Table 1 footnote b |
| MTP init   | `lf0`       | log(209000)    | CFU/mL | Clewe 2018 Table 1 (RSE 17%) |
| MTP init   | `ls0`       | log(324000)    | CFU/mL | Clewe 2018 Table 1 (RSE 12%) |
| RIF mono   | `lemax_fg_rif` | log(1) FIX   | 1/day  | Clewe 2018 Table 1 (Emax fixed at 1) |
| RIF mono   | `lec50_fg_rif` | log(0.388)   | mg/L   | Clewe 2018 Table 1 (RSE 19%) |
| RIF mono   | `gamma_fg_r`   | 2.8          | unitless | Clewe 2018 Table 1 (RSE 28%) |
| RIF mono   | `lemax_fd_rif` | log(1.97)    | 1/day  | Clewe 2018 Table 1 (RSE 3%) |
| RIF mono   | `lec50_fd_rif` | log(0.00303) | mg/L   | Clewe 2018 Table 1 (RSE 10%) |
| RIF mono   | `lemax_sd_rif` | log(1.79)    | 1/day  | Clewe 2018 Table 1 (RSE 4%) |
| RIF mono   | `lec50_sd_rif` | log(0.0113)  | mg/L   | Clewe 2018 Table 1 (RSE 32%) |
| RIF mono   | `lknd_rif`     | log(3.29)    | L/mg/day | Clewe 2018 Table 1 (RSE 17%) |
| INH mono   | `lemax_fd_inh` | log(22.2)    | 1/day  | Clewe 2018 Table 1 (RSE 35%) |
| INH mono   | `lec50_fd_inh` | log(0.168)   | mg/L   | Clewe 2018 Table 1 (RSE 34%) |
| INH mono   | `gamma_fd_h`   | 1.9          | unitless | Clewe 2018 Table 1 (RSE 11%) |
| INH mono   | `lemax_sd_inh` | log(8.55)    | 1/day  | Clewe 2018 Table 1 (RSE 17%) |
| INH mono   | `lec50_sd_inh` | log(0.0329)  | mg/L   | Clewe 2018 Table 1 (RSE 49%) |
| INH mono   | `gamma_sd_h`   | 1.74         | unitless | Clewe 2018 Table 1 (RSE 25%) |
| INH AR     | `lkon`         | log(0.0206)  | L/mg/day | Clewe 2018 Table 1 (RSE 31%) |
| INH AR     | `lkar_fd_inh`  | log(522)     | unitless | Clewe 2018 Table 1 (RSE 46%) |
| INH AR     | `lkar_sd_inh`  | log(2350)    | unitless | Clewe 2018 Table 1 (RSE 51%) |
| EMB mono   | `lemax_fd_emb` | log(2.21)    | 1/day  | Clewe 2018 Table 1 (RSE 1%) |
| EMB mono   | `lec50_fd_emb` | log(0.86)    | mg/L   | Clewe 2018 Table 1 (RSE 16%) |
| EMB mono   | `gamma_fd_e`   | 2.46         | unitless | Clewe 2018 Table 1 (RSE 23%) |
| EMB mono   | `lksd_emb`     | log(4.39)    | L/mg/day | Clewe 2018 Table 1 (RSE 69%) |
| GPDI F     | `int_fd_rif_inh` | -0.679     | fractional | Clewe 2018 Table 3 (RSE 11%) |
| GPDI F     | `int_fd_inh_rif` | 0 FIX      | fractional | Clewe 2018 Table 3 footnote b |
| GPDI F     | `int_fd_emb_inh` | 1.72       | fractional | Clewe 2018 Table 3 (RSE 15%) |
| GPDI F     | `int_fd_inh_emb` | 0 FIX      | fractional | Clewe 2018 Table 3 footnote b |
| GPDI F     | `int_fd_rif_emb` | -0.99 FIX  | fractional | Clewe 2018 Table 3 footnote c |
| GPDI F     | `int_fd_emb_rif` | -0.668     | fractional | Clewe 2018 Table 3 (RSE 22%) |
| GPDI S     | `int_sd_rif_inh` | 1.42       | fractional | Clewe 2018 Table 3 (RSE 22%) |
| GPDI S     | `int_sd_inh_rif` | 15.2       | fractional | Clewe 2018 Table 3 (RSE 49%) |
| GPDI S     | `int_sd_emb_inh` | 0.0963     | fractional | Clewe 2018 Table 3 (RSE 81%) |
| GPDI S     | `int_sd_inh_emb` | 164        | fractional | Clewe 2018 Table 3 (RSE 259%) |
| GPDI S     | `int_sd_rif_emb` | 486        | fractional | Clewe 2018 Table 3 (RSE 12%) |
| GPDI S     | `int_sd_emb_rif` | 2.09       | fractional | Clewe 2018 Table 3 (RSE 32%) |
| GPDI trio  | `int_sd_rif_inh_emb` | -0.749 | fractional | Clewe 2018 Table 3 (RSE 18%) |
| Residual   | `addSd`        | 1.0 FIX      | log(CFU/mL) | placeholder; not tabulated in Clewe 2018 (see Errata) |

Equation references:

* Adaptive resistance dynamics: Clewe 2018 Eqs 1-2 (with Eq 2 mass-balance
  correction noted in Errata).
* Bliss Independence combination of three drug effects on F and S: Clewe 2018
  Eq 8 (F bacterial sub-state) and Eq 9 (S bacterial sub-state).
* N-state ODE with only rifampicin acting: Clewe 2018 Eq 10.
* GPDI interaction term form: Clewe 2018 Table 3 footer ("The interaction
  parameters INT were estimated as 1 + INT_A,B * C_A / (EC50_A,B + C_A)") and
  Materials and methods, GPDI section (least complex baseline:
  EC50_A,B = EC50_A).
* MTP natural-growth transfer rates kFSlin, kSF, kFN, kSN, kNS fixed from the
  in vitro H37Rv MTP fit of Clewe 2016 JAC dkv416 (Clewe 2018 Table 1
  footnote b).
* Linear adaptive-resistance functional form (EC50_eff = EC50 * (1 + kAR *
  AR_on)): adapted from Mohamed 2014 JAC dkt520 for colistin /
  Pseudomonas aeruginosa adaptive resistance (Clewe 2018 reference 9).

# Virtual cohort

The simulations below reproduce the typical-value trajectories of Clewe 2018
Figure 1 (mean observed log10 cfu vs time) for natural growth and the three
mono-drug time-kill panels. Because drug concentrations are static (not PK
events), the cohort design uses a single dosing-free event table per regimen
and varies the three concentration covariates between simulations.

```{r cohort}
# Static observation grid spanning the 6-day in vitro experiment. The model
# carries five state ODEs (fbugs, sbugs, nbugs, aroff, aron) and three
# covariates (CONC_RIF_MGL, CONC_INH_MGL, CONC_EMB_MGL); there are no dosing
# events because the in vitro drug concentrations are constant by design.

obs_grid <- seq(0, 6, by = 0.05)

# One regimen helper -- returns a self-contained event table with the static
# drug concentrations encoded as covariate columns.
make_regimen <- function(id, regimen,
                         conc_rif = 0, conc_inh = 0, conc_emb = 0,
                         times = obs_grid) {
  tibble(
    id   = id,
    time = times,
    amt  = 0,
    evid = 0L,
    cmt  = "Cc",
    CONC_RIF_MGL = conc_rif,
    CONC_INH_MGL = conc_inh,
    CONC_EMB_MGL = conc_emb,
    regimen      = regimen
  )
}

# Concentrations replicated from Clewe 2018 Figure 1 panels.
events <- bind_rows(
  make_regimen( 1L, "Natural growth",         conc_rif = 0),
  make_regimen( 2L, "RIF 0.002 mg/L",         conc_rif = 0.002),
  make_regimen( 3L, "RIF 0.008 mg/L",         conc_rif = 0.008),
  make_regimen( 4L, "RIF 0.03 mg/L",          conc_rif = 0.03),
  make_regimen( 5L, "RIF 0.125 mg/L",         conc_rif = 0.125),
  make_regimen( 6L, "RIF 0.5 mg/L",           conc_rif = 0.5),
  make_regimen( 7L, "RIF 8 mg/L",             conc_rif = 8),
  make_regimen( 8L, "INH 0.01 mg/L",          conc_inh = 0.01),
  make_regimen( 9L, "INH 0.039 mg/L",         conc_inh = 0.039),
  make_regimen(10L, "INH 0.156 mg/L",         conc_inh = 0.156),
  make_regimen(11L, "INH 0.625 mg/L",         conc_inh = 0.625),
  make_regimen(12L, "INH 2.5 mg/L",           conc_inh = 2.5),
  make_regimen(13L, "INH 10 mg/L",            conc_inh = 10),
  make_regimen(14L, "INH 40 mg/L",            conc_inh = 40),
  make_regimen(15L, "EMB 0.0078 mg/L",        conc_emb = 0.0078),
  make_regimen(16L, "EMB 0.031 mg/L",         conc_emb = 0.031),
  make_regimen(17L, "EMB 0.125 mg/L",         conc_emb = 0.125),
  make_regimen(18L, "EMB 0.5 mg/L",           conc_emb = 0.5),
  make_regimen(19L, "EMB 2 mg/L",             conc_emb = 2),
  make_regimen(20L, "EMB 8 mg/L",             conc_emb = 8),
  make_regimen(21L, "EMB 32 mg/L",            conc_emb = 32),
  # Clinical-Cmax trio at last studied timepoint, used to check antagonism /
  # synergy classifications against Clewe 2018 Figure 5b.
  make_regimen(22L, "RIF 2 + INH 10 + EMB 8", conc_rif = 2, conc_inh = 10, conc_emb = 8)
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod <- readModelDb("Clewe_2018_TB_MTP_GPDI_invitro")

sim <- rxode2::rxSolve(
  mod, events = events,
  keep = c("regimen", "CONC_RIF_MGL", "CONC_INH_MGL", "CONC_EMB_MGL"),
  returnType = "data.frame"
) %>%
  as_tibble() %>%
  mutate(
    total_bugs = pmax(fbugs + sbugs + nbugs, 1),
    log10_total = log10(total_bugs)
  )
```

# Replicate published figures

## Figure 1 -- natural growth and rifampicin monotherapy

```{r figure-1-rif}
# Replicates Clewe 2018 Figure 1 top-left (natural growth) and top-right (RIF
# monotherapy at six concentrations).
rif_regimens <- c("Natural growth",
                  "RIF 0.002 mg/L", "RIF 0.008 mg/L", "RIF 0.03 mg/L",
                  "RIF 0.125 mg/L", "RIF 0.5 mg/L",   "RIF 8 mg/L")

sim %>%
  filter(regimen %in% rif_regimens) %>%
  mutate(regimen = factor(regimen, levels = rif_regimens)) %>%
  ggplot(aes(time, log10_total, colour = regimen)) +
  geom_line(linewidth = 0.7) +
  scale_y_continuous(name = "log10(CFU/mL)", limits = c(-0.2, 8)) +
  scale_x_continuous(name = "Time (days)", breaks = 0:6) +
  labs(title = "Clewe 2018 Figure 1 -- natural growth and rifampicin time-kill",
       caption = "Replicates Figure 1 top-left (open circles, natural growth) and top-right (RIF mono).") +
  theme_minimal() +
  theme(legend.title = element_blank())
```

## Figure 1 -- isoniazid monotherapy with adaptive resistance

```{r figure-1-inh}
# Replicates Clewe 2018 Figure 1 middle-left (INH mono, seven concentrations).
# At low INH (0.01-0.156 mg/L) the kill effect is weak and natural growth
# dominates. At intermediate INH (0.625-2.5 mg/L) the model shows initial kill
# followed by an apparent plateau / regrowth as the AR_on state grows and
# reduces INH potency. At high INH (10-40 mg/L) the kill is still meaningful
# despite adaptive resistance.
inh_regimens <- c("INH 0.01 mg/L", "INH 0.039 mg/L", "INH 0.156 mg/L",
                  "INH 0.625 mg/L", "INH 2.5 mg/L",  "INH 10 mg/L",  "INH 40 mg/L")

sim %>%
  filter(regimen %in% inh_regimens) %>%
  mutate(regimen = factor(regimen, levels = inh_regimens)) %>%
  ggplot(aes(time, log10_total, colour = regimen)) +
  geom_line(linewidth = 0.7) +
  scale_y_continuous(name = "log10(CFU/mL)", limits = c(-0.2, 8)) +
  scale_x_continuous(name = "Time (days)", breaks = 0:6) +
  labs(title = "Clewe 2018 Figure 1 -- isoniazid time-kill",
       caption = "Replicates Figure 1 middle-left. Adaptive-resistance loop modulates INH potency over time.") +
  theme_minimal() +
  theme(legend.title = element_blank())
```

## Figure 1 -- ethambutol monotherapy

```{r figure-1-emb}
# Replicates Clewe 2018 Figure 1 middle-right (EMB mono, seven concentrations).
emb_regimens <- c("EMB 0.0078 mg/L", "EMB 0.031 mg/L", "EMB 0.125 mg/L",
                  "EMB 0.5 mg/L",    "EMB 2 mg/L",     "EMB 8 mg/L",    "EMB 32 mg/L")

sim %>%
  filter(regimen %in% emb_regimens) %>%
  mutate(regimen = factor(regimen, levels = emb_regimens)) %>%
  ggplot(aes(time, log10_total, colour = regimen)) +
  geom_line(linewidth = 0.7) +
  scale_y_continuous(name = "log10(CFU/mL)", limits = c(-0.2, 8)) +
  scale_x_continuous(name = "Time (days)", breaks = 0:6) +
  labs(title = "Clewe 2018 Figure 1 -- ethambutol time-kill",
       caption = "Replicates Figure 1 middle-right. EMB acts as Emax on F-kill and linear on S-kill.") +
  theme_minimal() +
  theme(legend.title = element_blank())
```

# Validation checks (endogenous / mechanistic)

This is a typical-value mechanistic in vitro PD model; there is no
patient-level IIV and no PK structure. PKNCA-style NCA is not applicable.
The validation strategy follows the endogenous / mechanistic-model pattern
(`endogenous-validation.md` in the extraction skill): natural-growth
trajectory check, monotherapy potency check, conservation check for the
adaptive-resistance loop, and a combination antagonism / synergy check
against Figure 5b.

### 1. Natural-growth trajectory

With no drug, the F-state grows exponentially at rate kG = 0.796 1/day. The
S-state is drained by kSN to feed N and replenished by kFS(t) = kFSlin * t
from F; the N-state accumulates from F->N (slow) and S->N (faster). The total
CFU/mL at day 6 should be in the high 10^7 range, matching the open-circles
trajectory of Clewe 2018 Figure 1 top-left (terminal log10 ~ 7.5).

```{r natural-growth-check}
ng_traj <- sim %>%
  filter(regimen == "Natural growth", time %in% c(0, 1, 2, 3, 4, 5, 6)) %>%
  select(time, fbugs, sbugs, nbugs, total_bugs, log10_total)
knitr::kable(ng_traj, digits = 3, caption = "Natural-growth trajectory of F, S, N (CFU/mL).")
```

### 2. Rifampicin monotherapy potency check

Clewe 2018 Figure 1 top-right shows RIF 0.5 mg/L driving log10 CFU/mL to
roughly 1-2 by day 6, and RIF 8 mg/L driving log10 CFU/mL below the LOQ
(log10 = 0.7). The simulation should reproduce both qualitative endpoints
(deeper kill with higher RIF concentration; saturating at the LOQ).

```{r rif-potency-check}
sim %>%
  filter(regimen %in% c("Natural growth", "RIF 0.125 mg/L", "RIF 0.5 mg/L", "RIF 8 mg/L"),
         time == 6) %>%
  select(regimen, time, log10_total) %>%
  knitr::kable(digits = 3, caption = "Day-6 log10(CFU/mL) under increasing RIF monotherapy.")
```

### 3. Adaptive-resistance conservation

The AR_off and AR_on states represent the bacterial population fraction
susceptible vs adapted to isoniazid. With koff = 0 (no resistance reversal),
the kinetics are unidirectional: AR_off + AR_on must equal 1 at all times.
At any non-zero INH concentration, AR_on grows monotonically from 0 toward
1; in the limit of saturating INH for long enough, AR_on -> 1 and INH
becomes ineffective.

```{r ar-conservation-check}
sim %>%
  filter(regimen %in% c("Natural growth", "INH 0.625 mg/L", "INH 10 mg/L", "INH 40 mg/L"),
         time %in% c(0, 1, 3, 6)) %>%
  mutate(ar_total = aroff + aron) %>%
  select(regimen, time, aroff, aron, ar_total) %>%
  knitr::kable(digits = 4,
               caption = "aroff / aron / total over time -- total must equal 1.")
```

### 4. Trio combination at clinical Cmax (Figure 5b reference)

Clewe 2018 Figure 5b reports the model-predicted EBA (log10 cfu/mL/day) at
the clinically relevant unbound Cmax values RIF 2 mg/L, INH 10 mg/L, and
EMB 8 mg/L using the last studied timepoint (6 days). The published value
for the full triple combination is 1.34 log10 cfu/mL/day (classified as
antagonism relative to expected additivity). The simulation total log10
drop over 6 days corresponds to that EBA scaled by 6.

```{r trio-check}
sim %>%
  filter(regimen == "RIF 2 + INH 10 + EMB 8", time %in% c(0, 1, 3, 6)) %>%
  select(time, fbugs, sbugs, nbugs, log10_total) %>%
  knitr::kable(digits = 3,
               caption = "Clinical-Cmax trio (RIF 2 + INH 10 + EMB 8 mg/L) over the 6-day window.")
```

# Assumptions and deviations

* **GPDI functional form -- least-complex baseline EC50 anchors.** Clewe 2018
  Table 3 reports only INT magnitudes for the thirteen interactions; no
  EC50_INT parameters are tabulated, and the paper text states the "least
  complex" GPDI model was used as a starting point in which EC50_A,B = EC50_A
  (the modifier's mono EC50 for the same mechanism). The Materials and
  methods note that "Estimates of separate INT and interaction EC50
  parameters were then evaluated for statistical significance", but the
  Table 3 absence of EC50_INT values strongly suggests the least-complex
  baseline was retained for the final model. Implemented here: every
  modulation factor uses
  factor = 1 + INT_A,B * C_A / (EC50_A + C_A),
  where EC50_A is the modifier's mono EC50 for the same bacterial-state
  mechanism. When the modifier has no mechanism-specific EC50 (EMB has linear
  S-kill kSD_EMB; RIF has linear N-kill kND_RIF), the modifier's F-kill EC50
  is used as the saturation anchor. The Clewe 2018 Supplementary NONMEM model
  code (referenced in the paper but not on disk) would resolve this anchor
  choice unambiguously; users with access to that file should cross-check.
* **Linear adaptive resistance functional form.** Clewe 2018 Materials and
  methods presents Eq 3 (Emax form on isoniazid Emax) and Eq 4 (Emax form on
  isoniazid EC50) as the AR_max / AR_50 Hill-type adaptive-resistance
  alternatives evaluated, but Table 1 of the final model reports linear
  k_AR_FD_INH = 522 and k_AR_SD_INH = 2350 in place of (AR_max, AR_50)
  pairs. Implemented here as a linear modulation on EC50 (the published
  Discussion describes the resistance as affecting INH potency, not Emax):
  EC50_FD_INH_eff = EC50_FD_INH * (1 + k_AR_FD_INH * AR_on),
  EC50_SD_INH_eff = EC50_SD_INH * (1 + k_AR_SD_INH * AR_on).
  The units "(days^-1)" listed in Clewe 2018 Table 1 for k_AR are treated as
  a label slip because AR_on is dimensionless and the EC50 multiplier must be
  dimensionless. The supplement would confirm or refute this interpretation.
* **Mass-balance correction in adaptive-resistance Eq 2.** The published
  Clewe 2018 Eq 2 reads dAR_off/dt = -k_on * AR_off, but the conjugate Eq 1
  is dAR_on/dt = k_on * C_INH * AR_off. Without the C_INH factor on the
  outflow, AR_off + AR_on is not conserved. The model file encodes the
  mass-balanced form dAR_off/dt = -k_on * C_INH * AR_off so that AR_off +
  AR_on = 1 at all times. This is the form consistent with both Eq 1 and
  Materials and methods text ("rate of development of resistance ... was
  described using kon"), and the simulated AR_off + AR_on does indeed sum
  to 1 in the conservation check above.
* **kNF term in published Eq 8.** Clewe 2018 Eq 8 includes a "+ kNF * N"
  term that has no corresponding entry in Table 1 and no N->F arrow in
  Figure 2. The canonical Clewe 2016 MTP and the Chen 2017 mouse twin
  (modellib("Chen_2017_TB_MTP_GPDI_mouse")) omit any N->F transfer. This
  term is treated as a typographical inclusion and is not encoded; the F
  ODE in the model file matches the standard MTP structure
  dF/dt = kG * F * E_FG - kFS * F + kSF * S - kFN * F - E_FD * F.
* **Initial non-multiplying bacterial inoculum N0 = 0.** Clewe 2018
  Materials and methods does not list an initial N0; the standard MTP
  setup starts F0 and S0 from Table 1 and lets the N pool fill from
  F->N and S->N transfer. Encoded literally as `nbugs(0) <- 0`.
* **Static drug concentrations as covariates.** The in vitro experiment
  used fixed nominal drug concentrations over the 6-day window. The model
  encodes these as time-invariant covariates (`CONC_RIF_MGL`,
  `CONC_INH_MGL`, `CONC_EMB_MGL`) rather than as dosing events; the value
  is constant per simulation and varies between simulations. A future
  extension that adds a PK layer (e.g., for clinical EBA simulations) can
  reuse this PD layer with time-varying drug concentrations supplied by
  an upstream PK ODE -- see the in vivo twin
  modellib("Chen_2017_TB_MTP_GPDI_mouse") for the same MTP-GPDI PD layer
  driven by parallel one- and two-compartment PK ODEs.
* **Residual error placeholder.** Clewe 2018 does not tabulate the
  residual SD used in NONMEM (the M3-method censoring at the 5 CFU LOQ is
  described but the sigma value is absent from Table 1, Table 3, and the
  main text). The model file carries `addSd = fixed(1.0)` on the
  natural-log CFU/mL scale as a placeholder so the model is
  nlmixr2-fit-compatible; this is the same order of magnitude as the
  natural-log SD encoded by the Chen 2017 mouse twin (`sqrt(1.23) * log(10)
  ~ 2.55` on natural-log CFU/lungs after the log10-to-ln rescaling).
  Downstream users fitting this model to their own data should overwrite
  via `mod$ini` rather than relying on the placeholder.
* **No PD interaction on E_FG_RIF (RIF inhibition of F growth).** Clewe 2018
  Table 2 explicitly lists "none identified" for the PD interaction on
  E_FG_RIF. Implemented literally -- E_FG_RIF depends only on CONC_RIF_MGL
  and the four RIF / F-growth parameters (Emax_FG_RIF, EC50_FG_RIF,
  gamma_FG_R, kG).
* **No PD interaction on E_ND_RIF (RIF kill of N).** Clewe 2018 Table 2
  lists "none identified" for E_ND_RIF. Implemented literally -- the linear
  kND_RIF * CONC_RIF_MGL slope is the entire N-state kill effect.
* **Trio modulator anchor EC50.** The Clewe 2018 trio modulator
  INT^SD_RIF,INH|EMB = -0.749 applies to the
  INT^SD_RIF,INH * (1 + INT^SD_RIF,INH|EMB * C_EMB / (EC50_anchor + C_EMB))
  product. The paper provides no EC50_anchor value for the trio modulation;
  the model uses EC50_FD_EMB (= 0.86 mg/L) as the EMB saturation anchor
  because EMB has no S-kill EC50 (it is linear). The supplement would
  confirm this choice.
* **INT^FD_RIF,EMB encoded as -0.99 rather than -0.9999.** Clewe 2018
  Table 3 footnote c sets INT^FD_RIF,EMB to -0.9999 reflecting "a maximal
  decrease in EC50 identified in mono exposure". The model file encodes
  -0.99 to keep the multiplier (1 + INT) numerically well-conditioned
  (-0.9999 leaves a factor of 1e-4 in the denominator that can dominate
  rounding). The qualitative simulation behavior is unchanged at -0.99 vs
  -0.9999 within the saturating Emax term.
* **Non-canonical compartment names.** The model uses bacterial-state
  compartments `fbugs` / `sbugs` / `nbugs` and adaptive-resistance states
  `aroff` / `aron` (mirroring the Chen 2017 mouse twin). These do not
  match the canonical `depot` / `central` / `peripheral1` register and may
  trigger `checkModelConventions()` warnings; the deviation is intentional
  because the model carries no PK structure -- the three bacterial states
  and two resistance states are the entire state space.
* **Paper-specific drug-concentration covariates.** `CONC_RIF_MGL`,
  `CONC_INH_MGL`, and `CONC_EMB_MGL` are not in
  `inst/references/covariate-columns.md` because the canonical concentration
  concept in nlmixr2lib is a state-derived plasma concentration (`Cc`), not
  a static exogenous-drug-concentration covariate used to drive an in vitro
  PD model. These covariates are scoped to this specific model.

# Errata

* No erratum, corrigendum, or author correction was identified for the
  Clewe 2018 J Antimicrob Chemother article via the journal landing page
  (https://academic.oup.com/jac/article/73/2/437/4604762) or a PubMed
  "erratum" search as of 2026-05-22.
* Clewe 2018 Supplementary Data (containing Figures S1, S2 and the final
  MTP-GPDI NONMEM model code) is referenced in the paper but is not on
  disk in `from_people/literature_2018_search`. The model file extracts
  every parameter value verbatim from Table 1 (MTP + monotherapy effects)
  and Table 3 (GPDI interactions); the functional forms of the GPDI
  interactions, the linear adaptive-resistance form, and the mass-balance
  correction in Eq 2 are reconstructed from the canonical Wicha 2017 GPDI
  formulation, the Mohamed 2014 adaptive-resistance literature, and the
  Chen 2017 in vivo mouse twin (sharing senior authorship), and are
  documented above in Assumptions and deviations. A future operator with
  access to the Clewe 2018 Supplementary NONMEM file should cross-check
  the GPDI modulation-factor shape (Emax-style anchor, EC50 anchor for the
  trio modulator, and the linear vs Hill form of adaptive resistance).
* The published Eq 8 (F-state ODE) contains a "+ kNF * N" term with no
  Table 1 parameter and no corresponding arrow in Figure 2; treated here
  as a typographical inclusion (see Assumptions and deviations).
* The published Eq 2 (AR_off ODE) is missing the `C_INH` factor required
  for mass balance with Eq 1; the model file uses the conjugate
  mass-balanced form (see Assumptions and deviations).
* Table 1 lists the units of k_on and k_AR as "L*mg^-1*days" and "days^-1"
  respectively; both are treated as label slips. `k_on` (L*mg^-1*day^-1)
  multiplies `C_INH` (mg/L) and `AR_off` (unitless) to give a rate
  (1/day), and `k_AR` (dimensionless) multiplies `AR_on` (unitless) to
  give a dimensionless EC50 multiplier.
