---
title: "Clonidine in ICU adults (Cloesmeijer 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clonidine in ICU adults (Cloesmeijer 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Cloesmeijer_2020_clonidine"))$reference`
* Description: `r rxode2::rxode(readModelDb("Cloesmeijer_2020_clonidine"))$description`
* Article: <https://doi.org/10.1111/bcp.14273> (open access, PMC7373711)

Cloesmeijer et al. developed a two-compartment population PK model for
intravenous clonidine given as an add-on sedative to intubated adults in the
intensive care unit. The final model (Table 2 model 9, Table 3) has two
covariate relationships:

* central volume V1 is scaled to body weight with a fixed exponent of 1,
  `V1 = 124 L * (WT/70)`;
* clearance rises linearly with time after the start of the clonidine
  infusion, `CL = 17.0 L/h * (1 + 0.00213 * t)`, so the typical CL of 17.0 L/h
  at the start of treatment has risen to 20.5 L/h after 4 days.

IIV is on CL and V1 only, and the residual error is combined additive and
proportional.

# Population

Twenty-four intubated, sedated adults admitted to the ICU of Deventer Hospital
(The Netherlands; NCT02466373) with an expected stay of at least 3 days
received clonidine on top of standard sedation with morphine plus midazolam or
propofol. Eight patients each received a continuous infusion of 600, 1200 or
1800 ug/day (25, 50 or 75 ug/h); in each group four of the eight also received
a loading dose of 50% of the daily dose over 4 h. Table 1 reports 16 men and
8 women, median age 67 (25-83) years, weight 84 (53-113) kg, BMI 27 (20-44)
kg/m^2, serum creatinine 74 (32-441) umol/L and albumin 20 (12-32) g/L; three
patients received continuous veno-venous haemofiltration. The median treatment
duration was 96 (25-171) h. 275 plasma concentrations (5-16 per patient) were
analysed; the 11 below the 0.1 ug/L LLOQ were discarded in the final model.

The same information is available programmatically via
`rxode2::rxode(readModelDb("Cloesmeijer_2020_clonidine"))$population`.

# Source trace

Every `ini()` line in `inst/modeldb/specificDrugs/Cloesmeijer_2020_clonidine.R`
carries an in-file source-location comment. The table below collects them.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Two-compartment IV model, `d/dt(central)`, `d/dt(peripheral1)` | n/a | Results 3.1.1; Table 3 |
| Log-normal IIV, `theta_i = theta_pop * exp(eta_i)` | n/a | Methods Eq 1 |
| Combined additive + proportional residual error | n/a | Methods Eqs 2-3; Results 3.1.1 |
| V1 allometric scaling `(WT/70)^1`, exponent fixed | `e_wt_vc` = 1 (fixed) | Methods 2.3.2 Eq 4 and text; Results 3.1.3 |
| Linear effect of time after start of infusion on CL, `CL * (1 + slope * t)` | n/a | Methods Eq 6 and text; Table 2 model 9 |
| `lcl` (CL at the start of the infusion) | log(17.0 L/h) | Table 3 |
| `lvc` (V1 for a 70 kg patient) | log(124 L) | Table 3 |
| `lq` (Q) | log(83.7 L/h) | Table 3 |
| `lvp` (V2) | log(178 L) | Table 3 |
| `cl_time_slope` (fractional increase in CL per hour) | 0.00213 1/h | Table 3 'Increase CL per hour 0.213', stated as percent per hour in the Discussion |
| `etalcl` (IIV CL) | 33.3% CV -> 0.105189 | Table 3 |
| `etalvc` (IIV V1) | 66.8% CV -> 0.369689 | Table 3 |
| `propSd` | 0.141 (fraction) | Table 3 |
| `addSd` | 0.0532 ug/L | Table 3 |

# Virtual cohort

The observed data are not public. Section 2.5 of the paper describes the Monte
Carlo cohort behind Figure 4: body weight "followed a normal distribution from
53 to 113 kg with a mean of 84 kg". The standard deviation is not stated; the
cohort below uses 15 kg (a quarter of the range) and redraws any value outside
53-113 kg.

Two simulated designs are used:

* **Figure 4 regimens** (200 subjects per regimen, 4 days of infusion):
  (A) 50 ug/h; (B) 50 ug/h plus a 150 ug bolus given over 30 min at the start;
  (C) 100 ug/h for 6 h, then 50 ug/h.
* **The study design** (100 subjects per arm, six arms): 25, 50 or 75 ug/h
  for 96 h (the median treatment duration), with or without a preceding 4-h
  loading infusion of half the daily dose, followed by 48 h of washout.

```{r cohort}
# set.seed() fixes R's draws of body weight. It does not fix rxode2's
# simulation RNG across solver thread counts, so every assertion below is
# written to hold for any cohort the model can produce.
set.seed(20200514)

draw_wt <- function(n) {
  wt <- rnorm(n, mean = 84, sd = 15)
  while (any(out <- wt < 53 | wt > 113)) {
    wt[out] <- rnorm(sum(out), mean = 84, sd = 15)
  }
  wt
}

# One row per infusion segment for each regimen. `amt` is the total amount
# given in the segment and `rate` the ug/h infusion rate.
fig4_doses <- list(
  "A: 50 ug/h" = tibble(time = 0, amt = 50 * 96, rate = 50),
  "B: 50 ug/h + 150 ug over 30 min" = tibble(
    time = c(0, 0), amt = c(150, 50 * 96), rate = c(300, 50)
  ),
  "C: 100 ug/h for 6 h, then 50 ug/h" = tibble(
    time = c(0, 6), amt = c(100 * 6, 50 * 90), rate = c(100, 50)
  )
)

make_arm <- function(doses, label, n, id_offset, obs_times) {
  ids <- id_offset + seq_len(n)
  subj <- tibble(id = ids, WT = draw_wt(n))
  dose_rows <- tidyr::crossing(subj, doses) |>
    mutate(evid = 1L, cmt = "central")
  obs_rows <- tidyr::crossing(subj, time = obs_times) |>
    mutate(amt = NA_real_, rate = NA_real_, evid = 0L, cmt = "central")
  bind_rows(dose_rows, obs_rows) |>
    mutate(regimen = label) |>
    arrange(id, time, desc(evid))
}

fig4_times <- sort(unique(c(seq(0, 24, by = 0.25), seq(24, 96, by = 1))))
fig4_events <- bind_rows(lapply(seq_along(fig4_doses), function(i) {
  make_arm(fig4_doses[[i]], names(fig4_doses)[i],
    n = 200, id_offset = (i - 1) * 200, obs_times = fig4_times
  )
}))

# Study design: 96 h of maintenance infusion, optionally preceded by a 4-h
# loading infusion of 50% of the daily dose, then 48 h of washout.
study_doses <- list()
for (rate in c(25, 50, 75)) {
  study_doses[[sprintf("%d ug/h", rate)]] <-
    tibble(time = 0, amt = rate * 96, rate = rate)
  study_doses[[sprintf("%d ug/h + loading", rate)]] <- tibble(
    time = c(0, 4),
    amt = c(0.5 * rate * 24, rate * 92),
    rate = c(0.5 * rate * 24 / 4, rate)
  )
}
study_times <- sort(unique(c(seq(0, 12, by = 0.25), seq(12, 144, by = 1))))
study_events <- bind_rows(lapply(seq_along(study_doses), function(i) {
  make_arm(study_doses[[i]], names(study_doses)[i],
    n = 100, id_offset = 10000 + (i - 1) * 100, obs_times = study_times
  )
})) |>
  mutate(
    maint_rate = as.numeric(sub(" ug/h.*$", "", regimen)),
    loading = grepl("loading", regimen)
  )

stopifnot(
  length(unique(fig4_events$id)) == 600,
  length(unique(study_events$id)) == 600,
  !anyNA(fig4_events$WT),
  all(fig4_events$WT >= 53 & fig4_events$WT <= 113)
)
```

# Simulation

```{r simulate}
mod <- readModelDb("Cloesmeijer_2020_clonidine")

rxode2::rxSetSeed(20200514)
sim_fig4 <- rxode2::rxSolve(mod, events = fig4_events, keep = c("regimen", "WT")) |>
  as.data.frame()
sim_study <- rxode2::rxSolve(
  mod,
  events = study_events,
  keep = c("regimen", "WT", "maint_rate", "loading")
) |>
  as.data.frame()

stopifnot(!anyNA(sim_fig4$Cc), !anyNA(sim_study$Cc))
```

# Replicate Figure 4 -- dosing simulations

```{r figure-4, fig.height = 6, fig.cap = "Replicates Figure 4 of Cloesmeijer 2020: median and 5th-95th percentile clonidine concentrations (with residual error) for three regimens over 4 days. Dashed lines mark the 1.5-4.0 ug/L target range."}
fig4_summary <- sim_fig4 |>
  group_by(regimen, time) |>
  summarise(
    p05 = quantile(sim, 0.05),
    p50 = median(sim),
    p95 = quantile(sim, 0.95),
    .groups = "drop"
  )

ggplot(fig4_summary, aes(time, p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "grey80") +
  geom_line() +
  geom_hline(yintercept = c(1.5, 4), linetype = "dashed", colour = "grey40") +
  facet_wrap(~regimen, ncol = 1) +
  scale_x_continuous(breaks = seq(0, 96, 12)) +
  labs(x = "Time after start of infusion (h)", y = "Clonidine (ug/L)")
```

The paper quotes three numbers from these simulations: the time for 50% of the
population to reach 1.5 ug/L (14.5, 11 and 5 h for regimens A, B and C) and,
for regimen A, that 95% of patients are above 1.5 ug/L at steady state.

```{r figure-4-check}
time_to_target <- fig4_summary |>
  group_by(regimen) |>
  summarise(simulated_h = min(time[p50 >= 1.5]), .groups = "drop") |>
  mutate(published_h = c(14.5, 11, 5))

# Individual predictions (no residual error) for the steady-state claim.
# Clearance keeps rising, so the "steady state" is a slowly falling plateau;
# report the fraction above target on days 2, 3 and 4.
above_target <- sim_fig4 |>
  filter(regimen == names(fig4_doses)[1], time %in% c(48, 72, 96)) |>
  group_by(time) |>
  summarise(pct_above_1.5 = 100 * mean(ipredSim > 1.5), .groups = "drop")

time_to_target |>
  dplyr::rename(
    "Regimen" = regimen,
    "Simulated median time to 1.5 ug/L (h)" = simulated_h,
    "Published (h)" = published_h
  ) |>
  knitr::kable(caption = "Time for the median concentration to reach 1.5 ug/L (Results 3.2).")

above_target |>
  dplyr::rename(
    "Time after start (h)" = time,
    "Percent of subjects above 1.5 ug/L" = pct_above_1.5
  ) |>
  knitr::kable(
    digits = 1,
    caption = "Regimen A, percent of simulated subjects above 1.5 ug/L; published 95% at steady state.",
    row.names = FALSE
  )

stopifnot(
  nrow(time_to_target) == 3,
  # The median crossing is resolved on a 0.25 h grid and moves by about half
  # an hour between cohorts of 200; a CL or V1 transcription error moves it by
  # several hours.
  all(abs(time_to_target$simulated_h - time_to_target$published_h) < 2),
  # The model-true fraction is about 95% at 48 h and 94% at 96 h; with 200
  # subjects one binomial SE is about 1.6 points, so 88% sits more than four
  # SEs below. Halving the IIV or mis-scaling CL moves it well past this.
  nrow(above_target) == 3,
  all(above_target$pct_above_1.5 > 88)
)
```

The simulated median crossing times reproduce the published 14.5, 11 and 5 h,
and the fraction of patients above 1.5 ug/L on 1200 ug/day reproduces the
published 95%.

# Time-varying clearance

The Discussion states that the population CL "was 17 L/h at the start of the
treatment and increased to 20.4 L/h after 4 days". The typical-value
simulation returns the individual `cl`, so the claim can be checked directly.

```{r cl-check}
typical_events <- tibble(
  id = 1, time = c(0, 0, 96), amt = c(50 * 96, NA, NA), rate = c(50, NA, NA),
  evid = c(1L, 0L, 0L), cmt = "central", WT = 70
)
sim_typical <- rxode2::rxSolve(rxode2::zeroRe(mod), events = typical_events) |>
  as.data.frame()
cl_typ <- sim_typical$cl[sim_typical$time %in% c(0, 96)]
knitr::kable(
  tibble(
    "Time after start (h)" = c(0, 96),
    "Model CL (L/h)" = round(cl_typ, 2),
    "Published CL (L/h)" = c(17, 20.4)
  ),
  caption = "Typical clearance at the start of treatment and after 4 days (Discussion)."
)
# Deterministic: 17 * (1 + 0.00213 * 96) = 20.48, within 0.4% of the rounded
# 20.4 the paper prints.
stopifnot(
  abs(cl_typ[1] - 17) < 1e-6,
  abs(cl_typ[2] / 20.4 - 1) < 0.01
)
```

# Study-design simulation

Figure 3 of the paper is a visual predictive check against the observed data,
which are not available. The plot below shows the prediction intervals the
model gives for the study's six dosing arms, with a 48 h washout after 96 h of
infusion, as a visual check of the design the model was fitted to.

```{r figure-3, fig.height = 6, fig.cap = "Model-predicted median and 5th-95th percentile clonidine concentrations (with residual error) for the six arms of the Cloesmeijer 2020 study design; analogous to the visual predictive check of Figure 3 (observed data not available)."}
study_summary <- sim_study |>
  group_by(regimen, time) |>
  summarise(
    p05 = quantile(sim, 0.05),
    p50 = median(sim),
    p95 = quantile(sim, 0.95),
    .groups = "drop"
  ) |>
  mutate(regimen = factor(regimen, levels = names(study_doses)))

ggplot(study_summary, aes(time, p50)) +
  geom_ribbon(aes(ymin = pmax(p05, 0.01), ymax = p95), fill = "grey80") +
  geom_line() +
  geom_hline(yintercept = 0.1, linetype = "dotted", colour = "grey40") +
  facet_wrap(~regimen, ncol = 2) +
  scale_y_log10() +
  scale_x_continuous(breaks = seq(0, 144, 24)) +
  labs(
    x = "Time after start of infusion (h)", y = "Clonidine (ug/L, log scale)",
    caption = "Dotted line: 0.1 ug/L LLOQ."
  )
```

# PKNCA validation

The paper reports no NCA parameters, so PKNCA is used to check internal
consistency. Individual predictions are analysed over the 96 h infusion
(Cmax, AUC0-96) and over the 48 h washout (terminal half-life). The model is
linear in dose, so dose-normalised exposure must match across the 25, 50 and
75 ug/h arms.

```{r pknca}
conc_nca <- sim_study |>
  filter(!is.na(ipredSim)) |>
  transmute(id, time, conc = ipredSim, treatment = regimen)

conc_obj <- PKNCA::PKNCAconc(conc_nca, conc ~ time | treatment + id)
intervals <- data.frame(
  start = c(0, 96),
  end = c(96, 144),
  cmax = c(TRUE, FALSE),
  auclast = c(TRUE, FALSE),
  half.life = c(FALSE, TRUE)
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, intervals = intervals))
nca_df <- as.data.frame(nca_res)

nca_summary <- nca_df |>
  filter(PPTESTCD %in% c("cmax", "auclast", "half.life")) |>
  group_by(treatment, PPTESTCD) |>
  summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  mutate(treatment = factor(treatment, levels = names(study_doses))) |>
  arrange(treatment)

nca_summary |>
  dplyr::rename(
    "Arm" = treatment,
    "Cmax (ug/L)" = cmax,
    "AUC0-96 (h*ug/L)" = auclast,
    "Washout t1/2 (h)" = half.life
  ) |>
  knitr::kable(digits = 2, caption = "Median simulated NCA parameters per study arm (individual predictions).")
```

Two checks are deterministic and use the typical patient (70 kg, no random
effects), so they hold exactly on any machine:

* AUC0-96 divided by the maintenance rate is identical for the three arms
  without a loading dose (linearity);
* the washout half-life that PKNCA fits is close to the terminal half-life of
  the two-compartment system, `log(2) / beta`, evaluated at the clearance in
  force halfway through the washout (t = 120 h). Clearance keeps rising during
  the washout, so the match is approximate rather than exact.

```{r pknca-typical}
typ_events <- study_events |>
  filter(!loading) |>
  group_by(regimen) |>
  filter(id == min(id)) |>
  ungroup() |>
  mutate(WT = 70)
sim_typ <- rxode2::rxSolve(
  rxode2::zeroRe(mod),
  events = typ_events, keep = c("regimen", "maint_rate"),
  rtol = 1e-10, atol = 1e-12
) |>
  as.data.frame()

typ_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(
    sim_typ |> transmute(id, time, conc = Cc, treatment = regimen),
    conc ~ time | treatment + id
  ),
  intervals = intervals
)) |>
  as.data.frame() |>
  filter(PPTESTCD %in% c("auclast", "half.life")) |>
  select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  mutate(maint_rate = as.numeric(sub(" ug/h.*$", "", treatment)))

# Terminal half-life of the 2-compartment system at CL(120 h), V1 at 70 kg.
cl120 <- 17.0 * (1 + 0.00213 * 120)
k10 <- cl120 / 124
k12 <- 83.7 / 124
k21 <- 83.7 / 178
beta <- ((k10 + k12 + k21) - sqrt((k10 + k12 + k21)^2 - 4 * k10 * k21)) / 2
t_half_beta <- log(2) / beta

typ_nca |>
  mutate(
    auc_per_rate = auclast / maint_rate,
    beta_half_life = t_half_beta
  ) |>
  dplyr::rename(
    "Arm" = treatment,
    "AUC0-96 (h*ug/L)" = auclast,
    "AUC0-96 / rate (h^2/L)" = auc_per_rate,
    "PKNCA washout t1/2 (h)" = half.life,
    "log(2)/beta at CL(120 h) (h)" = beta_half_life
  ) |>
  select(-maint_rate) |>
  knitr::kable(digits = 3, caption = "Typical-patient NCA against the model's closed-form quantities.")

auc_per_rate <- typ_nca$auclast / typ_nca$maint_rate
stopifnot(
  nrow(typ_nca) == 3,
  # Linear kinetics: identical to integrator precision.
  max(abs(auc_per_rate / auc_per_rate[1] - 1)) < 1e-6,
  # Approximate: CL rises by about 9% across the 96-144 h washout.
  all(abs(typ_nca$half.life / t_half_beta - 1) < 0.1)
)
```

# Assumptions and deviations

* **Slope units.** Table 3 prints the time effect as "Increase CL per hour
  0.213" without a unit. The Discussion states "CL increased linearly with
  0.213%/h from baseline" and that the typical CL rises from 17 to 20.4 L/h
  over 4 days; `17 * (1 + 0.00213 * 96) = 20.5 L/h` confirms the percentage
  reading, so the slope is encoded as 0.00213 per hour.
* **Time origin.** `t` in the model is the time since the start of the
  clonidine infusion (Methods: "Cov was time in hours"), so the first dose must
  be at `t = 0`. The effect is not centred and has no plateau; it is valid over
  the studied treatment span (25-171 h) and extrapolation far beyond it is not
  supported by the data.
* **Allometry.** Methods Eq 4 mentions an allometric exponent of 0.75 for CL,
  but Results 3.1.3, the final model of Table 2 (model 9: "time after start
  infusion on CL + bodyweight on V1") and the Table 3 units (CL in L/h, V1 in
  L/70 kg) show that only V1 was weight-scaled in the final model. The model
  follows the final model.
* **Residual error.** Table 3 labels the proportional error "(%)" but prints
  0.141; both residual terms are read as standard deviations (14.1% and
  0.0532 ug/L). The RSE of 4% on the proportional term is at the information
  floor for an SD with 264 observations, which supports that reading.
* **IIV.** Table 3 reports %CV; the variances are `log(CV^2 + 1)`.
* **Virtual body weight.** Section 2.5 gives the Monte Carlo weight
  distribution as normal with mean 84 kg and range 53-113 kg but no SD; the
  vignette uses 15 kg and truncates to the range.
* **Loading dose timing.** The Figure 1 legend draws the loading dose and the
  maintenance infusion as consecutive segments, so the study-design simulation
  gives the 4-h loading infusion first and starts the maintenance rate at 4 h.
* **Screened covariates.** Creatinine clearance and albumin were significant
  on V1 in forward addition but dropped in backward elimination; they are
  recorded in `covariatesDataExcluded`. CVVH status was tested and not
  retained.
