---
title: "Midazolam and 1-OH-midazolam with efavirenz (Collins 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Midazolam and 1-OH-midazolam with efavirenz (Collins 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r load-model}
ui <- rxode2::rxode(readModelDb("Collins_2025_midazolam"))
```

# Model and source

* Citation: `r ui$reference`
* Article: <https://doi.org/10.1002/psp4.70116>
* Description: `r ui$description`

Collins 2025 used oral midazolam as a CYP3A probe to quantify how much
efavirenz induces CYP3A *in vivo*. Healthy volunteers received a 1 mg oral
midazolam dose on two occasions: on Day 1, one hour after a **single** 600 mg
efavirenz dose, and again on Day 24, after efavirenz 600 mg once daily on Days
7-23. Midazolam and 1-OH-midazolam were fitted jointly as a parent-metabolite
system, and the single-dose occasion is the model's reference state.

The design point that matters for interpreting every coefficient below is that
**both arms are efavirenz-exposed**. The contrast is induction-established
versus induction-not-yet-established, not efavirenz versus no efavirenz, which
is why the model uses the `CONMED_EFV_MD` covariate rather than `CONMED_EFV`.
The paper is explicit that this understates the full effect: "Our study design
did not include a midazolam-alone arm (without efavirenz), which would have
provided a baseline CYP3A activity for comparison ... the true magnitude of this
change may be underestimated."

# Population

```{r population}
pop <- ui$population
tibble::tibble(Field = names(pop),
               Value = vapply(pop, function(x) paste(
                 if (is.null(names(x))) x else paste0(names(x), " ", x, "%"),
                 collapse = "; "), character(1))) |>
  knitr::kable(caption = "Population metadata (Collins 2025 Table 1 and Methods).")
```

Seventy-two healthy volunteers (Table 1) completed the single-dose efavirenz
occasion; 58 of them also completed the multiple-dose occasion. Median age 25
years (18-50), median weight 73 kg (53-104), median BMI 24 kg/m^2 (18-32);
37.5% female; 72% White, 21% Black, 7% Other. Ninety-four percent were CYP3A4
normal metabolizers (*1/*1) and 75% were CYP3A5 non-expressers. The analysis
dataset held 2309 plasma concentrations (1153 midazolam, 1156
1-OH-midazolam). Plasma was assayed after beta-glucuronidase deconjugation, so
the 1-OH-midazolam measurements are total (free plus glucuronide) metabolite,
which is why metabolite concentrations run an order of magnitude above parent.

# Source trace

Every value in the model file comes from **Supporting Information File S5**, the
deposited final NONMEM control stream (`$PROBLEM Final`). The table below gives
the File S5 location, the value it implies, and the corresponding Table 2 entry.
Six rows disagree; the "Assumptions and deviations" section resolves them
against the paper's own Figure 4.

```{r source-trace}
tibble::tribble(
  ~Parameter,             ~`File S5 location`,        ~`Model value`,          ~`Table 2 (Final model estimate)`,
  "lka",                  "$THETA(1) = -0.303",       "ka = 0.7386 1/h",       "0.77  (disagrees)",
  "lcl",                  "$THETA(2) = 3.38",         "CL = 29.370 L/h",       "29.37",
  "lvc",                  "$THETA(3) = 1.67",         "Vc = 5.3122 L",         "5.31",
  "lvp",                  "$THETA(4) = 4.42",         "Vp = 83.096 L",         "83.10",
  "lq",                   "$THETA(5) = 2.73",         "Q = 15.333 L/h",        "15.33",
  "lcl_1ohm",             "$THETA(6) = 0.451",        "CLm = 1.5699 L/h",      "1.57",
  "lvc_1ohm",             "$THETA(7) = -0.416",       "Vcm = 0.65965 L",       "0.66",
  "lvp_1ohm",             "$THETA(8) = 3.02",         "Vpm = 20.492 L",        "20.49",
  "lq_1ohm",              "$THETA(9) = -0.494",       "Qm = 0.61019 L/h",      "0.61",
  "logitfdepot_mdefv",    "$THETA(10) = -0.511",      "F(MD) = expit = 0.3749", "0.60  (disagrees)",
  "e_conmed_efv_md_cl",   "$THETA(11) = 0.652",       "CL x 1.652",            "1.92  (disagrees)",
  "e_snp_cyp3a4_rs35599367_cl", "$THETA(12) = -0.117", "CL x 0.883",           "0.89  (disagrees)",
  "e_cyp3a5_expr_cl",     "$THETA(13) = 0.254",       "CL x 1.254",            "1.29  (disagrees)",
  "e_sexf_cl",            "$THETA(14) = 0.199",       "CL x 1.199",            "1.22  (disagrees)",
  "e_conmed_efv_md_ka",   "$THETA(15) = 0.255",       "ka x 1.255",            "1.29  (disagrees)",
  "lfdepot_sdefv",        "$PK F1 = 0.5",             "F(SD) = 0.50 fixed",    "0.50 (fixed)",
  "lfm",                  "$PK FMET = 0.7",           "fm = 0.70 fixed",       "Results text",
  "e_wt_cl_q",            "$PK LOG(WT/73)*(3/4)",     "0.75 fixed",            "Methods",
  "e_wt_vc_vp",           "$PK LOG(WT/73)",           "1 fixed",               "Methods",
  "etalka .. etalq_1ohm", "$OMEGA (9 variances)",     "log-scale variances",   "all 9 %CV match",
  "etalogitfdepot_mdefv", "$OMEGA = 0.134",           "logit-scale variance",  "37.9% CV matches",
  "propSd",               "$SIGMA = 0.0779",          "SD = 0.2791",           "27.9% CV",
  "propSd_1ohm",          "$SIGMA = 0.0688",          "SD = 0.2623",           "26.2% CV"
) |>
  knitr::kable(caption = "Source trace: File S5 (deposited final control stream) vs Table 2.")
```

The eight structural typical values, all ten omegas and both sigmas agree
exactly between File S5 and Table 2, which is what establishes File S5 as the
final model rather than a stale draft.

# Virtual cohort

```{r cohort}
MW_MIDAZOLAM <- 325.77                      # g/mol; 1 mg oral dose -> nmol
DOSE_NMOL    <- 1e-3 / MW_MIDAZOLAM * 1e9   # 3069.6 nmol

# Reference subject of the paper's Results sentence: 73 kg male, CYP3A4 *1/*1,
# CYP3A5 non-expresser.
REFERENCE_COVARIATES <- list(
  WT = 73, SEXF = 0, CYP3A5_EXPR = 0, SNP_CYP3A4_RS35599367 = 0
)

#' Build a single-subject event table: one oral midazolam dose into `depot`
#' plus observation rows on the `central` ODE state.
#'
#' The model declares two endpoints (`Cc` and `Cc_1ohm`), so observation rows
#' need a `dvid`. `dvid = 1L` is sufficient: rxode2 returns BOTH observables as
#' columns on every observation row, so one series yields both analytes.
make_events <- function(times, md, covariates = REFERENCE_COVARIATES, id = 1L) {
  dose <- data.frame(id = id, time = 0, amt = DOSE_NMOL, cmt = "depot",
                     evid = 1L, dvid = NA_integer_)
  obs  <- data.frame(id = id, time = times, amt = NA_real_, cmt = "central",
                     evid = 0L, dvid = 1L)
  ev <- rbind(dose, obs)
  for (nm in names(covariates)) ev[[nm]] <- covariates[[nm]]
  ev$CONMED_EFV_MD <- md
  ev[order(ev$time, -ev$evid), ]
}

#' Typical-value solve (IIV suppressed via omega = NA).
solve_typical <- function(events) {
  rxode2::rxSolve(ui, events, omega = NA, useLinCmt = FALSE,
                  keep = c("CONMED_EFV_MD", "WT", "SEXF", "CYP3A5_EXPR",
                           "SNP_CYP3A4_RS35599367"),
                  returnType = "data.frame")
}
```

`useLinCmt = FALSE` is required: rxode2's default ODE-to-`linCmt()`
auto-conversion corrupts the dvid mapping on multi-endpoint models.

# Replicate Figure 4

Figure 4 of Collins 2025 plots simulated midazolam (panel A) and
1-OH-midazolam (panel B) concentration-time curves after single-dose (black)
and multiple-dose (purple) efavirenz, over the study's 0.5-11 h sampling
window. This is the paper's only published quantitative simulation output, and
it is the evidence that decides the File S5 / Table 2 conflict.

```{r figure4}
FIG4_TIMES <- c(0.5, 1, 1.5, 2, 3, 5, 6, 7, 9, 11)

fig4 <- bind_rows(
  solve_typical(make_events(FIG4_TIMES, md = 0)),
  solve_typical(make_events(FIG4_TIMES, md = 1))
) |>
  mutate(occasion = ifelse(CONMED_EFV_MD == 1, "MD EFV", "SD EFV"))

fig4_long <- fig4 |>
  select(time, occasion, Midazolam = Cc, `1-OH-Midazolam` = Cc_1ohm) |>
  pivot_longer(c(Midazolam, `1-OH-Midazolam`), names_to = "analyte",
               values_to = "conc") |>
  mutate(analyte = factor(analyte, levels = c("Midazolam", "1-OH-Midazolam")))

ggplot(fig4_long, aes(time, conc, colour = occasion)) +
  geom_line(linewidth = 1) +
  facet_wrap(~analyte, scales = "free_y") +
  scale_y_log10() +
  scale_colour_manual(values = c("SD EFV" = "black", "MD EFV" = "purple")) +
  labs(x = "Time (hours)", y = "Plasma drug concentration (nM)", colour = NULL) +
  theme_bw()
```

Replicates Figure 4 of Collins 2025. Both analytes sit **lower** after
multiple-dose efavirenz. For midazolam that is expected (clearance is induced);
for the metabolite it is the diagnostic result, because total 1-OH-midazolam
exposure over all time is `fm * F * Dose / CLm` and depends on bioavailability
but **not** on parent clearance. A metabolite curve that falls under
multiple-dose efavirenz therefore says F went *down*.

## Scoring the two readings of the source against Figure 4

Panel A of the published figure was digitised from the article's own figure
file (three evenly-spaced decade labels give an exact 135.5 px/decade
calibration, so the parent-panel numbers below are accurate to the line
width). The MD/SD **ratio** is used rather than absolute concentrations because
the ratio is insensitive to whether the authors plotted a typical-value curve
or a median across simulated subjects.

```{r figure4-digitised}
# Operator-digitised from the Collins 2025 Figure 4 image file
# (psp470116 figure g005; panel A, log axis calibrated on the 10.0 / 1.0 / 0.1
# tick labels). Not a value published in numeric form by the authors.
fig4_digitised <- tibble::tribble(
  ~time, ~sd_obs, ~md_obs,
   0.5,  14.782,   8.582,
   1.0,  11.073,   5.708,
   1.5,   7.950,   3.669,
   2.0,   5.659,   2.359,
   3.0,   3.176,   1.126,
   5.0,   1.324,   0.383,
   6.0,   1.009,   0.289,
   7.0,   0.775,   0.220,
   9.0,   0.542,   0.157
) |>
  mutate(ratio_published = md_obs / sd_obs)

# Reading 1 (encoded): File S5 verbatim -- CL x 1.652, ka x 1.255, F = 0.3749.
ratio_fileS5 <- fig4 |>
  select(time, occasion, Cc) |>
  pivot_wider(names_from = occasion, values_from = Cc) |>
  transmute(time, ratio_fileS5 = `MD EFV` / `SD EFV`)

# Reading 2 (rejected): Table 2 back-transforms -- CL x 1.92, ka x 1.29,
# F = 0.60. Built by overriding the ini() values on the same model.
alt <- ui |>
  rxode2::ini(e_conmed_efv_md_cl = 0.92,   # 1.92 = Table 2
              e_conmed_efv_md_ka = 0.29,   # 1.29 = Table 2
              logitfdepot_mdefv = rxode2::logit(0.60))  # F = 0.60 = Table 2

ratio_table2 <- bind_rows(
  rxode2::rxSolve(alt, make_events(FIG4_TIMES, 0), omega = NA,
                  useLinCmt = FALSE, keep = "CONMED_EFV_MD",
                  returnType = "data.frame"),
  rxode2::rxSolve(alt, make_events(FIG4_TIMES, 1), omega = NA,
                  useLinCmt = FALSE, keep = "CONMED_EFV_MD",
                  returnType = "data.frame")
) |>
  mutate(occasion = ifelse(CONMED_EFV_MD == 1, "MD", "SD")) |>
  select(time, occasion, Cc) |>
  pivot_wider(names_from = occasion, values_from = Cc) |>
  transmute(time, ratio_table2 = MD / SD)

scoring <- fig4_digitised |>
  left_join(ratio_fileS5, by = "time") |>
  left_join(ratio_table2, by = "time") |>
  mutate(err_fileS5 = abs(log(ratio_fileS5 / ratio_published)),
         err_table2 = abs(log(ratio_table2 / ratio_published)))

scoring |>
  transmute(`Time (h)` = time,
            `Published MD/SD` = round(ratio_published, 3),
            `File S5 reading` = round(ratio_fileS5, 3),
            `Table 2 reading` = round(ratio_table2, 3)) |>
  knitr::kable(caption = paste(
    "Midazolam multiple-dose / single-dose concentration ratio: digitised from",
    "Collins 2025 Figure 4 panel A, versus the two readings of the source."))

fit_summary <- c(
  `File S5 (encoded)` = max(scoring$err_fileS5),
  `Table 2 back-transform` = max(scoring$err_table2)
)
round(fit_summary, 3)
```

```{r figure4-assert}
# The File S5 reading must reproduce every digitised ratio within 10% and must
# beat the Table 2 reading at every one of the nine time points.
stopifnot(
  max(scoring$err_fileS5) < log(1.10),
  all(scoring$err_fileS5 < scoring$err_table2)
)
```

The File S5 reading matches every digitised point to within
`r sprintf("%.1f%%", 100 * (exp(max(scoring$err_fileS5)) - 1))`; the Table 2
back-transform is off by up to
`r sprintf("%.0f%%", 100 * (exp(max(scoring$err_table2)) - 1))`, systematically
in one direction. The model file therefore encodes File S5.

# Structural checks

## The reference clearance the paper states in the text

Collins 2025 Results: "The clearance (CL/F) of midazolam for a 73 kg male with
normal CYP3A4 expression and reduced CYP3A5 expression after a single dose of
efavirenz was estimated at 29.37 L/h."

Clearance is read back out of the solved model rather than recomputed here, so
these checks test the model file itself and not a second transcription of it.

```{r reference-cl}
#' Typical-value clearance the model produces for a given covariate set.
cl_of <- function(md = 0, sexf = 0, expr = 0, s22 = 0, wt = 73) {
  ev <- make_events(1, md = md, covariates = list(
    WT = wt, SEXF = sexf, CYP3A5_EXPR = expr, SNP_CYP3A4_RS35599367 = s22))
  unique(solve_typical(ev)$cl)
}
stopifnot(abs(cl_of() - 29.37) < 0.005)
```

Note that despite the paper's "CL/F" wording this is a **true** clearance, not
an apparent one: File S5 codes bioavailability explicitly (`F1 = 0.5`), and
29.4 L/h is the expected value for systemic midazolam clearance in healthy
adults, whereas an oral CL/F at F = 0.5 would be near 59 L/h.

## Covariate multipliers

```{r covariate-multipliers}
tibble::tribble(
  ~Covariate,                       ~`File S5 THETA`, ~`Model multiplier`,   ~`Table 2`, ~`Figure 5 forest (bootstrap)`,
  "Multiple-dose efavirenz (CL)",   0.652,  cl_of(md = 1)   / cl_of(), 1.92, 1.92,
  "CYP3A4*22 carrier (CL)",        -0.117,  cl_of(s22 = 1)  / cl_of(), 0.89, 0.940,
  "CYP3A5 expresser (CL)",          0.254,  cl_of(expr = 1) / cl_of(), 1.29, 1.27,
  "Female (CL)",                    0.199,  cl_of(sexf = 1) / cl_of(), 1.22, 1.30
) |>
  mutate(`Model multiplier` = round(`Model multiplier`, 3)) |>
  knitr::kable(caption = paste(
    "Covariate multipliers on midazolam clearance. The model uses the File S5",
    "coded (1 + THETA) form; Table 2 and Figure 5 report exp(THETA) and the",
    "bootstrap median respectively."))

stopifnot(
  abs(cl_of(md   = 1) / cl_of() - 1.652) < 1e-9,
  abs(cl_of(s22  = 1) / cl_of() - 0.883) < 1e-9,
  abs(cl_of(expr = 1) / cl_of() - 1.254) < 1e-9,
  abs(cl_of(sexf = 1) / cl_of() - 1.199) < 1e-9,
  # Allometry: doubling weight multiplies CL by 2^0.75.
  abs(cl_of(wt = 146) / cl_of() - 2^0.75) < 1e-9
)
```

## Bioavailability carries between-subject variability on one occasion only

File S5 puts `ETA(10)` inside the `IF(EFV.EQ.0)` branch, so single-dose
bioavailability is a fixed 0.5 with no IIV while the multiple-dose value is
drawn per subject. This is unusual enough to be worth asserting directly.

`rxSolve()` returns the model's derived `fdepot` as a column, so this can be
checked exactly rather than through a sampling statistic.

```{r f-iiv}
set.seed(20250321)
N_F <- 200
f_cohort <- function(md) {
  do.call(rbind, lapply(seq_len(N_F), function(i) make_events(1, md = md, id = i)))
}
f_sd <- rxode2::rxSolve(ui, f_cohort(0), useLinCmt = FALSE,
                        returnType = "data.frame")$fdepot
f_md <- rxode2::rxSolve(ui, f_cohort(1), useLinCmt = FALSE,
                        returnType = "data.frame")$fdepot

# Typical values, IIV suppressed.
f_typ <- vapply(c(0, 1), function(md) unique(solve_typical(make_events(1, md))$fdepot),
                numeric(1))

tibble::tibble(
  Occasion            = c("SD EFV", "MD EFV"),
  `Typical F`         = round(f_typ, 4),
  `Distinct F values` = c(length(unique(f_sd)), length(unique(f_md))),
  `F range`           = c(sprintf("%.4f", unique(f_sd)),
                          sprintf("%.3f - %.3f", min(f_md), max(f_md)))
) |>
  knitr::kable(caption = paste(
    "Bioavailability across", N_F, "simulated subjects per occasion."))

stopifnot(
  # Single-dose F is a fixed 0.5 for every subject: no between-subject spread.
  all(f_sd == 0.5),
  length(unique(f_sd)) == 1L,
  # Multiple-dose F is drawn per subject and stays inside (0, 1) by construction
  # of the logit transform.
  length(unique(f_md)) == N_F,
  all(f_md > 0 & f_md < 1),
  # Typical values are exactly the File S5 constants.
  abs(f_typ[1] - 0.5) < 1e-12,
  abs(f_typ[2] - 1 / (1 + exp(0.511))) < 1e-9
)
```

# PKNCA validation

The paper reports no non-compartmental analysis, so there is no published NCA
table to compare against. Instead the NCA is scored against the model's own
closed-form exposure identities, which are exact for a linear system:

* parent: `AUC(0-inf) = F * Dose / CL`
* metabolite: `AUC(0-inf) = fm * F * Dose / CLm` -- independent of parent
  clearance, because every molecule of midazolam that is cleared has already
  had its chance to become 1-OH-midazolam.

```{r pknca}
NCA_GRID <- sort(unique(c(seq(0, 12, by = 0.05), seq(12, 48, by = 0.25),
                          seq(48, 240, by = 1))))

nca_sim <- bind_rows(
  solve_typical(make_events(NCA_GRID, md = 0)),
  solve_typical(make_events(NCA_GRID, md = 1))
) |>
  mutate(treatment = ifelse(CONMED_EFV_MD == 1, "MD EFV", "SD EFV"),
         id = 1L)

run_nca <- function(conc_col, concu) {
  sim_nca <- nca_sim |>
    dplyr::filter(!is.na(.data[[conc_col]])) |>
    transmute(id, time, Cc = .data[[conc_col]], treatment)

  # Guarantee a time-zero row so PKNCA does not warn about an AUC range
  # starting before the first measurement (pre-dose oral concentration is 0).
  sim_nca <- bind_rows(
    sim_nca,
    sim_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
  ) |>
    distinct(id, treatment, time, .keep_all = TRUE) |>
    arrange(treatment, id, time)

  dose_df <- data.frame(id = 1L, time = 0, amt = DOSE_NMOL,
                        treatment = c("SD EFV", "MD EFV"))

  conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                               concu = concu, timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                               doseu = "nmol")
  intervals <- data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                          aucinf.obs = TRUE, half.life = TRUE)
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}

nca_parent <- run_nca("Cc", "nM")
nca_metab  <- run_nca("Cc_1ohm", "nM")
```

```{r pknca-table}
F_SD <- 0.5
F_MD <- 1 / (1 + exp(0.511))          # expit(-0.511)
CL_SD <- 29.370; CL_MD <- CL_SD * 1.652
CLM   <- 1.5699; FM <- 0.7

expected <- tibble::tibble(
  treatment  = c("SD EFV", "MD EFV"),
  parent_auc = c(F_SD * DOSE_NMOL / CL_SD, F_MD * DOSE_NMOL / CL_MD),
  metab_auc  = c(FM * F_SD * DOSE_NMOL / CLM, FM * F_MD * DOSE_NMOL / CLM)
)

grab <- function(res, code) {
  as.data.frame(res$result) |>
    dplyr::filter(PPTESTCD == code) |>
    dplyr::select(treatment, value = PPORRES)
}

nca_tbl <- expected |>
  left_join(grab(nca_parent, "aucinf.obs") |> rename(parent_nca = value),
            by = "treatment") |>
  left_join(grab(nca_metab, "aucinf.obs") |> rename(metab_nca = value),
            by = "treatment") |>
  left_join(grab(nca_parent, "cmax") |> rename(parent_cmax = value),
            by = "treatment") |>
  left_join(grab(nca_parent, "half.life") |> rename(parent_thalf = value),
            by = "treatment")

nca_tbl |>
  transmute(
    Occasion                      = treatment,
    `Midazolam Cmax (nM)`         = round(parent_cmax, 2),
    `Midazolam t1/2 (h)`          = round(parent_thalf, 2),
    `Midazolam AUCinf, NCA`       = round(parent_nca, 2),
    `Midazolam AUCinf, F*Dose/CL` = round(parent_auc, 2),
    `1-OH AUCinf, NCA`            = round(metab_nca, 1),
    `1-OH AUCinf, fm*F*Dose/CLm`  = round(metab_auc, 1)
  ) |>
  knitr::kable(caption = paste(
    "PKNCA output versus the model's closed-form exposure identities.",
    "AUC in nM*h."))
```

```{r pknca-assert}
# Per-occasion identity: PKNCA AUCinf must reproduce the closed form to 1%.
stopifnot(
  all(abs(nca_tbl$parent_nca / nca_tbl$parent_auc - 1) < 0.01),
  all(abs(nca_tbl$metab_nca  / nca_tbl$metab_auc  - 1) < 0.01)
)

# Metabolite exposure is set by bioavailability alone, so its MD/SD ratio must
# equal F_MD / F_SD exactly -- the identity that makes Figure 4 panel B a
# bioavailability read-out.
metab_ratio <- nca_tbl$metab_nca[nca_tbl$treatment == "MD EFV"] /
  nca_tbl$metab_nca[nca_tbl$treatment == "SD EFV"]
stopifnot(abs(metab_ratio - F_MD / F_SD) < 0.005)

# Parent exposure falls by the induction factor divided by the F change.
parent_ratio <- nca_tbl$parent_nca[nca_tbl$treatment == "MD EFV"] /
  nca_tbl$parent_nca[nca_tbl$treatment == "SD EFV"]
stopifnot(abs(parent_ratio - (F_MD / F_SD) / 1.652) < 0.005)
```

```{r pknca-report, echo = FALSE}
cat(sprintf(
  "Metabolite AUC ratio (MD/SD) = %.3f, equals F_MD/F_SD = %.3f\nParent AUC ratio (MD/SD) = %.3f\n",
  metab_ratio, F_MD / F_SD, parent_ratio))
```

Multiple-dose efavirenz lowers total 1-OH-midazolam exposure by 25%, exactly
tracking the drop in bioavailability from 0.50 to 0.375, and lowers midazolam
exposure by 55% -- the combined effect of a 65% clearance increase and the 25%
bioavailability decrease.

# Between-subject variability

```{r vpc}
set.seed(70116)
N_PER_ARM <- 100     # 100 subjects per efavirenz occasion
VPC_TIMES <- c(0.5, 1, 1.5, 2, 3, 5, 6, 7, 9, 11)

vpc_cohort <- function(md, id_offset) {
  wt <- stats::runif(N_PER_ARM, 53, 104)          # Table 1 weight range
  sexf <- stats::rbinom(N_PER_ARM, 1, 0.375)      # Table 1: 37.5% female
  expr <- stats::rbinom(N_PER_ARM, 1, 0.25)       # Table 1: 25% CYP3A5 expressers
  s22 <- stats::rbinom(N_PER_ARM, 1, 0.06)        # Table 1: 6% CYP3A4 IM
  do.call(rbind, lapply(seq_len(N_PER_ARM), function(i)
    make_events(VPC_TIMES, md = md, id = id_offset + i,
                covariates = list(WT = wt[i], SEXF = sexf[i],
                                  CYP3A5_EXPR = expr[i],
                                  SNP_CYP3A4_RS35599367 = s22[i]))))
}

vpc <- bind_rows(
  rxode2::rxSolve(ui, vpc_cohort(0, 0), useLinCmt = FALSE,
                  keep = "CONMED_EFV_MD", returnType = "data.frame"),
  rxode2::rxSolve(ui, vpc_cohort(1, 1000), useLinCmt = FALSE,
                  keep = "CONMED_EFV_MD", returnType = "data.frame")
) |>
  mutate(occasion = ifelse(CONMED_EFV_MD == 1, "MD EFV", "SD EFV"))

vpc_long <- vpc |>
  select(time, occasion, Midazolam = Cc, `1-OH-Midazolam` = Cc_1ohm) |>
  pivot_longer(c(Midazolam, `1-OH-Midazolam`), names_to = "analyte",
               values_to = "conc") |>
  mutate(analyte = factor(analyte, levels = c("Midazolam", "1-OH-Midazolam"))) |>
  group_by(analyte, occasion, time) |>
  summarise(p05 = quantile(conc, 0.05), p50 = median(conc),
            p95 = quantile(conc, 0.95), .groups = "drop")

ggplot(vpc_long, aes(time, p50, colour = occasion, fill = occasion)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 1) +
  facet_wrap(~analyte, scales = "free_y") +
  scale_y_log10() +
  scale_colour_manual(values = c("SD EFV" = "black", "MD EFV" = "purple")) +
  scale_fill_manual(values = c("SD EFV" = "black", "MD EFV" = "purple")) +
  labs(x = "Time (hours)", y = "Plasma drug concentration (nM)",
       colour = NULL, fill = NULL) +
  theme_bw()
```

Median and 5th-95th percentile individual predictions for 100 simulated
subjects per occasion, with covariates drawn from the Table 1 marginal
distributions. Compare with Figure 3 of Collins 2025 (visual predictive checks
stratified by single- and multiple-dose efavirenz). The very wide midazolam
band reflects the 117% CV on central volume reported in Table 2, which
dominates the early distribution phase.

# Assumptions and deviations

## 1. Parameter values are taken from File S5, not from Table 2

This is the one substantive decision in the extraction, and it changes six
numbers, including the paper's headline result.

The deposited final control stream (Supporting Information File S5) and Table 2
agree exactly on all eight structural typical values, all ten omegas and both
sigmas -- 20 of 21 comparable quantities. They disagree on the five covariate
coefficients and the multiple-dose bioavailability, and the disagreement has a
single, coherent explanation: **Table 2 reports `exp(THETA)` for every THETA**,
which is the correct back-transformation for the MU-referenced log-scale
typical values but the wrong one for the six parameters that File S5 codes
differently.

| File S5 coded form | THETA | Correct value | Table 2 = exp(THETA) |
|---|---|---|---|
| `(1 + L0*THETA(11))` on CL | 0.652 | 1.652 | 1.92 |
| `(1 + M1*THETA(12))` on CL | -0.117 | 0.883 | 0.89 |
| `(1 + N1*THETA(13))` on CL | 0.254 | 1.254 | 1.29 |
| `(1 + K0*THETA(14))` on CL | 0.199 | 1.199 | 1.22 |
| `(1 + L0*THETA(15))` on Ka | 0.255 | 1.255 | 1.29 |
| `EXP(LF1)/(1+EXP(LF1))` on F | -0.511 | 0.3749 | 0.60 |

The "Replicate Figure 4" section above scores both readings against the
paper's own simulated profiles and the File S5 reading wins at every time
point, by a wide margin. Two independent strands support it:

* **The metabolite direction.** Total 1-OH-midazolam exposure is
  `fm * F * Dose / CLm` and does not involve parent clearance, so the
  metabolite panel of Figure 4 is a direct read-out of bioavailability. The
  published metabolite curve is clearly *lower* after multiple-dose efavirenz,
  which requires `F < 0.5`. Table 2's `F = 0.60` would put it about 60% higher
  than the single-dose curve.
* **The parent ratio.** The digitised midazolam MD/SD ratio runs 0.58 at 0.5 h
  down to 0.28 in the tail; the File S5 reading predicts 0.60 to 0.29, the
  Table 2 reading 0.54 to 0.22.

Consequences for anyone reading the paper alongside this model:

* The **abstract, Discussion, Conclusion and Figure 5 forest plot all state
  that multiple-dose efavirenz raises midazolam clearance 1.92-fold.** The
  deposited control stream gives **1.652-fold**. The paper's own Study
  Highlights section independently states "1.64-fold", which agrees with File
  S5 and not with the abstract.
* The paper's Discussion devotes a paragraph to explaining why bioavailability
  appeared to *increase* with chronic efavirenz ("The reason for this
  observation remains unclear ... we speculate that the apparent increase in
  oral midazolam bioavailability ... reflects a scenario where intestinal
  CYP3A remains uninduced"). Under File S5, bioavailability **decreased**, from
  0.50 to 0.375. That removes the puzzle entirely and is consistent with the
  Kharasch et al. result the paper itself cites, in which alfentanil
  bioavailability fell after 14 days of efavirenz.
* The abstract's covariate values (1.27 CYP3A5, 0.94 CYP3A4, 1.30 female) are
  the **bootstrap medians** from Table 2's third column, not the final-model
  estimates, so they differ from Table 2's second column as well.

This has not been raised with the authors and no erratum was found; a reader
who needs the published numbers rather than the deposited ones can recover them
by overriding the six `ini()` entries as the scoring chunk above does.

## 2. Table 2's Ka does not reconcile with File S5 under either reading

`THETA(1) = -0.303` gives `ka = exp(-0.303) = 0.7386 1/h`, but Table 2 reports
0.77 (bootstrap median also 0.77, 95% CI 0.71-0.84). `ka` is a plain
MU-referenced log-scale typical value, so `exp()` is the right transform here
and the two sources simply disagree by 4%. This is the only quantity that
neither hypothesis explains. The model uses File S5 for internal consistency;
0.7386 lies inside the published bootstrap interval, and the difference is
immaterial to exposure.

## 3. Values digitised from a figure

The nine midazolam concentration pairs in the scoring chunk were read off the
Collins 2025 Figure 4 image file by the extractor; they are not published in
numeric form. They are used only to adjudicate between two published parameter
sets, never to fit or tune anything. The panel A log axis carries three
evenly-spaced decade labels, giving an exact pixel-per-decade calibration.

## 4. Covariate encoding

* `CONMED_EFV_MD` is a newly registered canonical covariate for this
  extraction. Both of its levels are efavirenz-exposed, so it is deliberately
  *not* `CONMED_EFV` (which contrasts efavirenz against a non-efavirenz
  comparator) and *not* `MULTI_DOSE_PT` (which flags the multiple-dose phase of
  the analyte, whereas here the analyte is single-dosed on both occasions and
  the perpetrator's dosing history is what changes).
* Two source columns are polarity-inverted relative to their canonical form and
  are transformed in `covariateData`: `SEXF = 1 - SEX` (the paper's `SEX` is
  0 = female) and `CONMED_EFV_MD = 1 - EFV` (the paper's `EFV` is 0 = multiple
  dose). Both inversions are stated explicitly in the paper's Results text and
  in File S5's inline comments, so neither is inferred.
* The paper's "CYP3A4 intermediate metabolizer" stratum is exactly the
  `SNP_CYP3A4_RS35599367` carrier stratum: genotyping was for *22 (rs35599367)
  and the cohort contained only *1/*1 and *1/*22 subjects.

## 5. Not represented in the model

* **CYP2B6 genotype** was assayed but deliberately excluded by the authors
  ("Incorporation of CYP2B6 genotype was deemed unnecessary, given the low
  frequency of poor metabolizers in the study cohort"), so it is not a
  covariate here.
* **Race, age and BMI** were collected (Table 1) but not carried into the final
  model; they are documented in `population` rather than `covariateData`.
* **Occasion structure.** 58 of the 72 subjects contribute both efavirenz
  occasions, and File S5 has no inter-occasion variability terms -- the
  disposition etas are subject-level and shared across a subject's two
  occasions. Simulating an occasion at a time, as this vignette does, matches
  the fitted model; simulating a crossover would additionally require holding
  each subject's etas fixed across their two records.
* **Below-quantification-limit data.** 33% of midazolam concentrations were
  below the 0.3 ng/mL limit of quantification and were retained uncensored
  under the Keizer "all data" approach. A simulation model has no BLQ, so
  nothing is encoded; but the parent parameters -- particularly the 117% CV on
  central volume -- carry the influence of that choice.
* **The other three cocktail probes** (caffeine, tolbutamide, omeprazole) were
  co-administered but are outside this model.
```
