---
title: "Zavegepant (Comisar 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Zavegepant (Comisar 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Comisar_2025_zavegepant"))$reference`
* Description: `r rxode2::rxode(readModelDb("Comisar_2025_zavegepant"))$description`
* Article: <https://doi.org/10.1002/psp4.13257>

Zavegepant (ZAVZPRET) is a small-molecule calcitonin gene-related peptide (CGRP)
receptor antagonist and the first CGRP antagonist approved as a nasal spray for
acute treatment of migraine in adults. Comisar 2025 pooled nine phase I studies
in which zavegepant was given intravenously, intranasally, or orally, and fit a
three-compartment model with first-order elimination and sequential zero-order
then first-order absorption. A tenth study (BHV3500-106) was held out for
external validation.

# Population

The model-development dataset comprised 277 participants contributing 10,819
quantifiable plasma concentrations from nine phase I studies conducted in North
America (Comisar 2025 Table 1, Table 2, and Results "Data summary"). Of the
12,253 concentration records screened across all 594 individuals enrolled, 1434
(11.7%) were excluded -- 226 (1.8%) collected before the first zavegepant dose
and 1208 (9.9%) below the 0.4 ng/mL lower limit of quantitation.

Median age was 40 years (range 18-71) and median body weight 74 kg (range
49-131). 126/277 (45.5%) were female, 245 (88.4%) Caucasian and 32 (11.6%)
Black, with no Asian or Other participants; 159 (57.4%) were Hispanic/Latino.
Most were healthy volunteers: 39 (14.1%) were patients with migraine and 8
(2.9%) had moderate (Child-Pugh score 7-9) hepatic impairment. Median creatinine
clearance was 115 mL/min (range 53-211); the Discussion notes that only 17
participants had mild and one had moderate renal impairment, so the renal
screen was underpowered at the impaired end. By route, 209 participants (75.5%)
received intranasal zavegepant, 62 (22.4%) oral soft gelatin capsules and 6
(2.2%) an intravenous infusion.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Comisar_2025_zavegepant")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Comisar_2025_zavegepant.R`. The table below
collects them in one place for review. All parameter values come from Comisar
2025 Table 3 ("Parameter estimates for the final population pharmacokinetic
model"); the structure comes from Figure 1 and the Methods / Results narrative.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` | 13.3 L/h | Table 3, "Elimination clearance, CL" (RSE 9.9%) |
| `lvc` | 12.1 L | Table 3, "Central volume of distribution, V2" (RSE 8.8%) |
| `lq` | 2.6 L/h | Table 3, "Intercompartmental clearance ... first peripheral compartment, Q3" (RSE 10.5%) |
| `lvp` | 66.0 L | Table 3, "First peripheral volume of distribution, V3" (RSE 10.7%) |
| `lq2` | 5.2 L/h | Table 3, "Intercompartmental clearance ... second peripheral compartment, Q4" (RSE 16.3%) |
| `lvp2` | 10.7 L | Table 3, "Second peripheral volume of distribution, V4" (RSE 10.1%) |
| `lfdepot_intranasal` | 5.1% | Table 3, "Bioavailability of intranasal zavegepant" (RSE 11.0%) |
| `lfdepot_oral` | 0.65% | Table 3, "Bioavailability of oral zavegepant" (RSE 17.5%) |
| `lka_intranasal` | 5.8 1/h | Table 3, "First-order absorption rate constant (ka) for intranasal zavegepant" (RSE 5.1%) |
| `lka_oral` | 0.81 1/h | Table 3, "First-order absorption rate constant (ka) for oral zavegepant" (RSE 6.1%) |
| `ld1_intranasal` | 8.6 min = 0.1433 h | Table 3, "Duration of zero-order absorption for intranasal zavegepant" (RSE 8.7%); quoted as 0.14 h in Results |
| `ld1_oral` | 57.2 min = 0.9533 h | Table 3, "Duration of zero-order absorption for oral zavegepant" (RSE 14.4%); quoted as 0.95 h in Results |
| `ltlag_oral` | 12.2 min = 0.2033 h | Table 3, "Absorption lag for oral zavegepant" (RSE 4.3%) |
| `e_wt_cl` | 0.75 (fixed) | Methods "Population pharmacokinetic modeling"; Table 3 Note |
| `e_wt_vc` | 1 (fixed) | Methods "Population pharmacokinetic modeling"; Table 3 Note |
| `e_hepimp_mod_cl` | -43.9% | Table 3, "Moderate hepatic impairment on CL" (RSE 13.5%) |
| `e_conmed_rif_cl` | -41.1% | Table 3, "Co-administration of rifampin on CL" (RSE 12.1%) |
| `e_conmed_itraconazole_cl` | -27.9% | Table 3, "Co-administration of itraconazole on clearance when zavegepant administered orally" (RSE 26.3%) |
| `e_fed_fdepot_oral` | -50.9% | Table 3, "Food effect on bioavailability of oral zavegepant" (RSE 9.8%) |
| `etalfdepot` | 66.1% IIV | Table 3 bioavailability rows (RSE 4.8%, shrinkage 6.8%) |
| `etalka` | 44.0% IIV | Table 3 ka rows (RSE 11.1%, shrinkage 41.1%) |
| `etald1` | 115% IIV | Table 3 D1 rows (RSE 7.7%, shrinkage 11.6%) |
| `etalvc` | 32.2% IIV | Table 3 V2 row (RSE 10.9%, shrinkage 28%) |
| `etalvp` | 26.1% IIV | Table 3 V3 row (RSE 23.1%, shrinkage 49.9%) |
| `etalcl` | 26.8% IIV | Table 3 CL row (RSE 6.1%, shrinkage 17.8%) |
| `propSd` | 37.8% | Table 3, "Proportional error" (RSE 2.9%) |
| `addSd` | 0.001 (fixed) | Table 3, "Additive error (FIXED)", printed as ng/L |
| Three-compartment ODE system, first-order elimination from central | n/a | Figure 1 schematic; Results "Population pharmacokinetic model development" |
| Sequential zero-order (`dur(depot)`) then first-order (`ka`) absorption | n/a | Figure 1 caption ("D1, zero-order infusion duration into the first (depot) compartment"); Methods "Population pharmacokinetic modeling" |
| Allometric scaling on all disposition parameters, reference 70 kg | n/a | Methods; Results "Covariate analysis" ("tested and applied to all disposition parameters") |
| Fractional `(1 + theta * X)` categorical covariate form | n/a | Methods "Population pharmacokinetic modeling" ("theta_x is the proportional constant of the covariate effect model") |

# Virtual cohort

Original observed data are not publicly available. The simulations below use
virtual populations whose covariate distributions approximate the published
trial demographics (Comisar 2025 Table 2): log-normal body weight with median
74 kg truncated to the observed 49-131 kg range.

Five arms are simulated, each with 100 participants (well inside the 200 per arm
cap):

* `IV 10 mg` -- intravenous bolus into `central`, the reference arm that fixes
  the true (not apparent) CL and V values reported in Table 3.
* `IN 10 mg` -- 10 mg intranasal, the approved migraine dose.
* `IN 10 mg hepatic` -- 10 mg intranasal in moderate hepatic impairment, the
  comparator for Figure 2.
* `PO 50 mg fasted` and `PO 50 mg fed` -- oral soft gelatin capsule, the
  comparator for the food effect on oral bioavailability.

```{r cohort}
set.seed(20250112)

n_per_arm <- 100L

# Weight distribution matching Comisar 2025 Table 2: median 74 kg, range 49-131.
draw_wt <- function(n) {
  wt <- rlnorm(n, meanlog = log(74), sdlog = 0.20)
  pmin(pmax(wt, 49), 131)
}

# Build one arm as a self-contained event table. `id_offset` shifts subject IDs
# so arms can be bind_rows()-ed without colliding; duplicate IDs across arms are
# silently merged by rxSolve into a single subject receiving the summed dose.
make_arm <- function(label, dose_mg, cmt_dose, route_oral, fed = 0,
                     hepimp = 0, id_offset = 0L) {
  subj <- tibble(
    id                  = id_offset + seq_len(n_per_arm),
    treatment           = label,
    WT                  = draw_wt(n_per_arm),
    ROUTE_ORAL          = route_oral,
    FED                 = fed,
    HEPIMP_MOD          = hepimp,
    CONMED_RIF          = 0,
    CONMED_ITRACONAZOLE = 0
  )

  # rate = -2 tells rxode2 to apply the model's dur(depot) zero-order window
  # rather than treating the extravascular dose as an instantaneous bolus.
  # The IV arm doses straight into `central` and takes a plain bolus (rate = 0).
  dosing <- subj |>
    mutate(
      time = 0,
      amt  = dose_mg,
      evid = 1L,
      rate = if (identical(cmt_dose, "depot")) -2 else 0,
      cmt  = cmt_dose
    )

  # Dense early grid to resolve the intranasal Tmax (ka = 5.8 1/h, D1 = 8.6 min),
  # thinning out into the long terminal phase driven by the 66 L compartment.
  obs_times <- unique(c(
    seq(0, 2, by = 0.05),
    seq(2, 12, by = 0.25),
    seq(12, 72, by = 1),
    seq(72, 168, by = 4)
  ))

  obs <- subj |>
    crossing(time = obs_times) |>
    mutate(
      amt  = NA_real_,
      evid = 0L,
      rate = 0,
      # `central` is an ODE state; Cc is an algebraic observable and must NOT be
      # used as a cmt (that would auto-inject a slot and renumber the states).
      cmt  = "central"
    )

  bind_rows(dosing, obs) |> arrange(id, time, desc(evid))
}

events <- bind_rows(
  make_arm("IV 10 mg",         10, "central", route_oral = 0, id_offset =   0L),
  make_arm("IN 10 mg",         10, "depot",   route_oral = 0, id_offset = 100L),
  make_arm("IN 10 mg hepatic", 10, "depot",   route_oral = 0, hepimp = 1,
           id_offset = 200L),
  make_arm("PO 50 mg fasted",  50, "depot",   route_oral = 1, fed = 0,
           id_offset = 300L),
  make_arm("PO 50 mg fed",     50, "depot",   route_oral = 1, fed = 1,
           id_offset = 400L)
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod <- readModelDb("Comisar_2025_zavegepant")

sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep   = c("treatment", "WT", "ROUTE_ORAL", "FED", "HEPIMP_MOD")
) |>
  as.data.frame()

stopifnot(dplyr::n_distinct(sim$id) == 5L * n_per_arm)
stopifnot(all(sim$Cc >= 0, na.rm = TRUE))
```

`Cc` is the individual prediction (`IPRED`) without residual error; the `sim`
column returned by `rxSolve` carries the additive-plus-proportional residual
error on top of it.

# Replicate published results

## Concentration-time profiles by route

Comisar 2025 Figure 3b shows prediction-corrected visual predictive checks
stratified by route of administration (intravenous, nasal spray, oral soft
gelatin capsule). The panels below are the corresponding simulated 5th, 50th and
95th percentiles from the packaged model.

```{r figure-3b}
sim |>
  filter(treatment %in% c("IV 10 mg", "IN 10 mg", "PO 50 mg fasted"),
         time <= 48) |>
  group_by(treatment, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~treatment, scales = "free_y") +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Zavegepant plasma concentration (ng/mL)",
    title = "Simulated concentration-time profiles by route",
    caption = "Compare with Figure 3b of Comisar 2025 (VPC stratified by route)."
  )
```

## Allometric scaling (Discussion)

Comisar 2025 Discussion reports the population mean parameter ratios relative to
the typical 70 kg individual at the 5th-percentile (57 kg) and highest (95 kg)
body weights: clearance 85.7% and 126%, volume of distribution 81.4% and 136%.
These are exact consequences of the fixed exponents and are reproduced here from
the packaged model, not from the printed ratios.

```{r allometry}
allo_events <- tibble(
  id = 1:3, WT = c(57, 70, 95), ROUTE_ORAL = 0, FED = 0, HEPIMP_MOD = 0,
  CONMED_RIF = 0, CONMED_ITRACONAZOLE = 0
) |>
  mutate(time = 0, amt = 10, evid = 1L, rate = 0, cmt = "central") |>
  bind_rows(
    tibble(
      id = 1:3, WT = c(57, 70, 95), ROUTE_ORAL = 0, FED = 0, HEPIMP_MOD = 0,
      CONMED_RIF = 0, CONMED_ITRACONAZOLE = 0
    ) |>
      crossing(time = c(0.5, 1)) |>
      mutate(amt = NA_real_, evid = 0L, rate = 0, cmt = "central")
  ) |>
  arrange(id, time, desc(evid))

allo <- rxode2::rxSolve(rxode2::zeroRe(mod), events = allo_events,
                        keep = "WT") |>
  as.data.frame() |>
  group_by(WT) |>
  summarise(cl = first(cl), vc = first(vc), vp = first(vp), vp2 = first(vp2),
            .groups = "drop") |>
  mutate(vss = vc + vp + vp2)

ref <- allo |> filter(WT == 70)

allo_tab <- allo |>
  transmute(
    `Body weight (kg)`   = WT,
    `CL (L/h)`           = round(cl, 2),
    `CL vs 70 kg (%)`    = round(100 * cl / ref$cl, 1),
    `Vss (L)`            = round(vss, 1),
    `Vss vs 70 kg (%)`   = round(100 * vss / ref$vss, 1)
  )

knitr::kable(
  allo_tab,
  caption = paste(
    "Allometric scaling of the typical subject. Comisar 2025 Discussion reports",
    "CL 85.7% at 57 kg and 126% at 95 kg, and volume 81.4% and 136%."
  )
)

# The published ratios are quoted to three significant figures, so match to
# 0.05 percentage points rather than merely "close".
stopifnot(
  abs(allo_tab$`CL vs 70 kg (%)`[allo_tab$`Body weight (kg)` == 57] - 85.7) < 0.05,
  abs(allo_tab$`CL vs 70 kg (%)`[allo_tab$`Body weight (kg)` == 95] - 125.7) < 0.05,
  abs(allo_tab$`Vss vs 70 kg (%)`[allo_tab$`Body weight (kg)` == 57] - 81.4) < 0.05,
  abs(allo_tab$`Vss vs 70 kg (%)`[allo_tab$`Body weight (kg)` == 95] - 135.7) < 0.05
)

# Typical Vss at 70 kg must be V2 + V3 + V4 = 12.1 + 66.0 + 10.7 = 88.8 L
# (Comisar 2025 Discussion).
stopifnot(abs(ref$vss - 88.8) < 1e-6)
```

The 95 kg CL ratio lands at 125.7%, which is the exact value of
`(95/70)^0.75`; the paper rounds it to 126%. The 95 kg volume ratio is 135.7%
against the paper's 136%. Both are rounding, not disagreement.

## Moderate hepatic impairment (Figure 2)

Comisar 2025 Figure 2 shows the distribution of relative mean exposure
(AUC0-24 and Cmax) after a single 10 mg intranasal dose in participants with
moderate hepatic impairment versus the reference population, with predicted mean
increases of 63% and 12% respectively. The ratios below are typical-value
predictions from the packaged model; the paper's are medians of a distribution
generated by sampling 3000 parameter sets from the variance-covariance matrix,
so they are the comparable central estimates but not identical constructions.

```{r figure-2}
hep_events <- bind_rows(
  make_arm("Normal hepatic function", 10, "depot", route_oral = 0,
           hepimp = 0, id_offset = 0L),
  make_arm("Moderate hepatic impairment", 10, "depot", route_oral = 0,
           hepimp = 1, id_offset = 100L)
) |>
  # Typical-value comparison: hold weight at the allometric reference so the
  # only difference between the arms is HEPIMP_MOD.
  mutate(WT = 70)

hep <- rxode2::rxSolve(rxode2::zeroRe(mod), events = hep_events,
                       keep = c("treatment", "HEPIMP_MOD")) |>
  as.data.frame()

hep_exposure <- hep |>
  filter(time <= 24) |>
  group_by(treatment) |>
  arrange(time, .by_group = TRUE) |>
  summarise(
    cmax     = max(Cc),
    auc_0_24 = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    .groups  = "drop"
  )

hep_ratio <- tibble(
  Metric = c("AUC0-24", "Cmax"),
  `Simulated ratio (impaired / normal)` = c(
    hep_exposure$auc_0_24[hep_exposure$treatment == "Moderate hepatic impairment"] /
      hep_exposure$auc_0_24[hep_exposure$treatment == "Normal hepatic function"],
    hep_exposure$cmax[hep_exposure$treatment == "Moderate hepatic impairment"] /
      hep_exposure$cmax[hep_exposure$treatment == "Normal hepatic function"]
  ),
  `Comisar 2025 reported increase` = c("+63%", "+12%")
) |>
  mutate(`Simulated increase` = sprintf("%+.0f%%", 100 * (`Simulated ratio (impaired / normal)` - 1)),
         `Simulated ratio (impaired / normal)` = round(`Simulated ratio (impaired / normal)`, 3))

knitr::kable(
  hep_ratio,
  caption = paste(
    "Relative exposure in moderate hepatic impairment after a single 10 mg",
    "intranasal dose. Compare with Figure 2 of Comisar 2025."
  )
)

# The CL reduction is -43.9%, so the untruncated AUC ratio would be
# 1 / 0.561 = 1.783. AUC0-24 is truncated well before the terminal phase and
# must therefore sit strictly between 1 and that bound, near the paper's 1.63.
auc_ratio <- hep_ratio$`Simulated ratio (impaired / normal)`[hep_ratio$Metric == "AUC0-24"]
cmax_ratio <- hep_ratio$`Simulated ratio (impaired / normal)`[hep_ratio$Metric == "Cmax"]
stopifnot(auc_ratio > 1, auc_ratio < 1 / (1 - 0.439))
stopifnot(abs(auc_ratio - 1.63) < 0.15)
# Cmax is absorption-limited for a low-clearance compound given intranasally, so
# the clearance effect barely moves it.
stopifnot(cmax_ratio > 1, cmax_ratio < 1.25)
```

## Food effect on oral bioavailability

Table 3 reports a -50.9% food effect on oral bioavailability, so the fed/fasted
exposure ratio must be exactly 0.491 for otherwise-identical subjects. Because
both oral arms were drawn with the same seed offsets but different random
weights, this check is run on typical-value subjects.

```{r food-effect}
food_events <- bind_rows(
  make_arm("PO 50 mg fasted", 50, "depot", route_oral = 1, fed = 0,
           id_offset = 0L),
  make_arm("PO 50 mg fed",    50, "depot", route_oral = 1, fed = 1,
           id_offset = 100L)
) |>
  mutate(WT = 70)

food <- rxode2::rxSolve(rxode2::zeroRe(mod), events = food_events,
                        keep = c("treatment", "FED")) |>
  as.data.frame()

food_ratio <- food |>
  group_by(treatment) |>
  arrange(time, .by_group = TRUE) |>
  summarise(auc = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
            .groups = "drop")

fed_fasted <- food_ratio$auc[food_ratio$treatment == "PO 50 mg fed"] /
  food_ratio$auc[food_ratio$treatment == "PO 50 mg fasted"]

# Bioavailability scales AUC linearly, so the ratio is the multiplier exactly.
stopifnot(abs(fed_fasted - (1 - 0.509)) < 1e-3)
cat(sprintf("Fed / fasted oral AUC ratio: %.4f (expected %.4f)\n",
            fed_fasted, 1 - 0.509))
```

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  # Only !is.na(Cc): a `time > 0` or `Cc > 0` filter would drop the time-zero
  # row that PKNCA needs to anchor AUC.
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)

# Guarantee a time = 0 row per (id, treatment).
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- events |>
  filter(evid == 1) |>
  select(id, time, amt, treatment)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- suppressWarnings(PKNCA::pk.nca(nca_data))

nca_wide <- as.data.frame(nca_res) |>
  select(treatment, id, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

nca_summary <- nca_wide |>
  group_by(treatment) |>
  summarise(
    n            = dplyr::n(),
    cmax_med     = median(cmax, na.rm = TRUE),
    tmax_med     = median(tmax, na.rm = TRUE),
    aucinf_med   = median(aucinf.obs, na.rm = TRUE),
    halflife_med = median(half.life, na.rm = TRUE),
    .groups      = "drop"
  )

nca_summary |>
  mutate(across(c(cmax_med, aucinf_med, halflife_med), \(x) signif(x, 3)),
         tmax_med = round(tmax_med, 2)) |>
  rename(
    "Arm"                       = treatment,
    "N"                         = n,
    "Cmax (ng/mL)"              = cmax_med,
    "Tmax (h)"                  = tmax_med,
    "AUC0-inf (ng*h/mL)"        = aucinf_med,
    "t1/2 (h)"                  = halflife_med
  ) |>
  knitr::kable(caption = "Median simulated NCA parameters by arm (PKNCA).")
```

## Per-subject mass-balance identity

Comisar 2025 reports no NCA parameter table, so there is no published Cmax /
AUC to compare against directly. The stronger available check is the exact
per-subject identity `AUC0-inf = F * Dose / CL`, which must hold for every
individual in a linear model. Because `fdepot` and `cl` are returned as columns
by `rxSolve`, this is testable without any tolerance on the population median
(a median over 100 draws carries several percent of noise; the per-subject
identity carries none).

```{r auc-identity}
per_subject <- sim |>
  group_by(id, treatment) |>
  summarise(cl = first(cl), fdepot = first(fdepot), .groups = "drop") |>
  left_join(
    events |> filter(evid == 1) |> select(id, amt),
    by = "id"
  ) |>
  left_join(nca_wide |> select(id, treatment, aucinf.obs),
            by = c("id", "treatment")) |>
  mutate(
    # IV doses go straight into `central`, so F = 1 for that arm; every other
    # arm passes through the depot and carries the route-specific F.
    f_applied = ifelse(treatment == "IV 10 mg", 1, fdepot),
    # amt is mg and cl is L/h, so amt/cl is mg*h/L; x1000 gives ng*h/mL.
    auc_expected = 1000 * f_applied * amt / cl,
    rel_error    = aucinf.obs / auc_expected - 1
  )

identity_tab <- per_subject |>
  group_by(treatment) |>
  summarise(
    n                    = dplyr::n(),
    `max |rel. error|`   = max(abs(rel_error), na.rm = TRUE),
    .groups              = "drop"
  )

identity_tab |>
  mutate(`max |rel. error|` = signif(`max |rel. error|`, 3)) |>
  rename("Arm" = treatment, "N" = n) |>
  knitr::kable(
    caption = paste(
      "Per-subject agreement between PKNCA AUC0-inf and F * Dose / CL.",
      "Residual error is numerical (observation-grid resolution and the",
      "log-linear extrapolation of the terminal tail), not structural."
    )
  )

# Fail loudly if the identity breaks for ANY subject in ANY arm. 2% accommodates
# the trapezoidal grid and terminal extrapolation only.
stopifnot(nrow(per_subject) == 5L * n_per_arm)
stopifnot(!anyNA(per_subject$rel_error))
stopifnot(max(abs(per_subject$rel_error)) < 0.02)
```

## Comparison against published derived quantities

Comisar 2025 does not tabulate NCA results, but the Discussion derives several
secondary quantities from the model. Those are reproduced here from the packaged
model.

```{r derived-quantities}
typ <- allo |> filter(WT == 70)
f_in <- exp(rxode2::rxode(mod)$theta[["lfdepot_intranasal"]])

derived <- tibble::tribble(
  ~Quantity,                                   ~Simulated,                         ~`Comisar 2025`,
  "Vss at 70 kg (L)",                          round(typ$vss, 1),                  88.8,
  "CL at 70 kg (L/h)",                         round(typ$cl, 1),                   13.3,
  "Apparent CL of intranasal zavegepant (L/h)", round(typ$cl / f_in, 0),           266,
  "Apparent Vss of intranasal zavegepant (L)",  round(typ$vss / f_in, 0),          1774
)

knitr::kable(
  derived,
  caption = paste(
    "Derived quantities from Comisar 2025 Discussion. The two apparent-parameter",
    "rows disagree by ~2% because the paper divided by an unrounded F of about",
    "5.0% while Table 3 prints 5.1%; see Assumptions and deviations."
  )
)

# Vss and CL are exact; the apparent quantities inherit the paper's own
# F-rounding inconsistency and are checked only to that tolerance.
stopifnot(abs(typ$vss - 88.8) < 0.05, abs(typ$cl - 13.3) < 0.05)
stopifnot(abs(typ$cl / f_in / 266 - 1) < 0.03)
stopifnot(abs(typ$vss / f_in / 1774 - 1) < 0.03)
```

# Assumptions and deviations

* **IIV convention.** Comisar 2025 Table 3 reports inter-individual variability
  as "%IIV" with no footnote naming the convention. The values are read here as
  log-normal CV%, so `omega^2 = log(CV^2 + 1)` (the convention documented in the
  nlmixr2lib parameter-naming reference). The alternative reading
  `omega^2 = CV^2` would give variances of 0.4369 (F), 0.1936 (ka), 1.3225 (D1),
  0.1037 (Vc), 0.0681 (Vp) and 0.0718 (CL) instead of the values used. The two
  readings are close for every term except D1, whose 115% IIV maps to
  `omega = 0.918` under the convention used here versus `omega = 1.150` under
  the alternative. All typical-value results in this vignette are unaffected;
  only the width of the simulated variability bands changes.
* **Shared etas across routes.** Each "%IIV" cell in Table 3 spans both the
  intranasal and the oral row of the quantity it belongs to, and the Methods
  list the evaluated IIV terms one-per-quantity rather than one-per-route
  ("bioavailability of intranasal and soft gelatin capsules", etc.). A single
  eta is therefore shared between routes for F, ka and D1. Table 3 reports no
  off-diagonal covariances, so all etas are simulated independently.
* **Minutes converted to hours.** D1 and the oral absorption lag are tabulated
  in minutes while the rest of the model is on an hours time base. They are
  divided by 60 here, which is what the paper does itself when it quotes
  "a D1 of 0.14 h" and "a D1 of 0.95 h" in the Results. The exact minute values
  are used rather than the paper's two-significant-figure hour roundings.
* **Route encoding.** `ROUTE_ORAL` was registered as a new canonical covariate
  column under the `ROUTE_*` family policy in
  `inst/references/covariate-columns.md`. Intravenous doses are distinguished
  from intranasal ones by the dose record's `cmt` (they are placed directly into
  `central`) rather than by a second indicator column, so a single binary
  covariate covers the three-route design.
* **Rifampin route scope.** Table 3 restricts the itraconazole clearance effect
  to oral zavegepant ("Co-administration of itraconazole on clearance when
  zavegepant administered orally") but places no route restriction on the
  adjacent rifampin row, and the Discussion preserves exactly the same contrast
  -- itraconazole "was not a significant covariate for intranasal zavegepant
  clearance, but decreased oral zavegepant clearance by 27.9%", against a plain
  "Co-administration of rifampin ... reduced zavegepant clearance by 41.1%". The
  rifampin effect is therefore encoded here as route-independent.

  Table 1 explains both why the two rows differ and why the choice is
  numerically moot. Study BHV3500-111 evaluated "the effects of rifampin on the
  pharmacokinetics of **oral** zavegepant and ... the effects of itraconazole on
  the pharmacokinetics of zavegepant administered **orally or as a nasal
  spray**". The dataset therefore contains no intranasal-plus-rifampin
  observations at all: the authors had no route contrast available to test for
  rifampin, so no restriction appears on that row, whereas for itraconazole they
  did have one, tested it, and found the effect only for oral. Because
  `CONMED_RIF` is never set on an intranasal or intravenous record in a dataset
  resembling the paper's own, the route-independent and oral-restricted
  encodings give identical predictions there; they diverge only if a user
  deliberately simulates a combination the study never observed.

  The paper's prose is nonetheless inconsistent about this in three places. The
  Abstract says moderate hepatic impairment or rifampin "decreased elimination
  clearance of **oral** zavegepant by ~40%"; the Results covariate paragraph
  compresses the two DDIs into "co-administration of rifampin and itraconazole
  on CL of **orally administered** zavegepant (-41.1%, -27.9%)"; and the Study
  Highlights box says the same effects decreased clearance of "zavegepant
  **nasal spray** by ~40%". The Abstract sentence is demonstrably unreliable on
  this point because it applies the same "oral" qualifier to moderate hepatic
  impairment, which the Discussion identifies as "the only clinically relevant
  covariate found for **intranasal** zavegepant administration" and which
  Figure 2 evaluates entirely on a 10 mg intranasal dose. Table 3 is the
  parameter source and is unambiguous.
* **Additive residual error units.** Table 3 prints the fixed additive error as
  "0.001 ng/L" although the assay and the LLOQ (0.4 ng/mL) are in ng/mL. This is
  almost certainly a typo for ng/mL. Under either reading the term is a
  numerically negligible stabilising placeholder at least five orders of
  magnitude below the LLOQ, so it is entered as the printed 0.001 on the model's
  ng/mL concentration scale and has no visible effect on any simulation.
* **Apparent-parameter arithmetic.** The Discussion states that apparent
  intranasal clearance is "266 L/h (calculated by CL/intranasal bioavailability
  = 13.3 L/5.1%)", but 13.3/0.051 = 260.8, not 266. Likewise 88.8/0.051 = 1741,
  not the quoted 1774. Both printed values correspond to dividing by an
  unrounded bioavailability of about 5.00%, so the underlying estimate was
  slightly below the 5.1% printed in Table 3. The tabulated 5.1% is used as the
  parameter here, and the derived-quantities check above is tolerated to 3%
  accordingly.
* **Female percentage conflict.** Table 2 reports 126/277 (45.5%) female while
  the Results narrative states "female (49.3%)". The Table 2 value is used in
  the model's `population` metadata because it carries the underlying count.
* **Figure 1 compartment labels.** The Figure 1 caption labels the two
  intercompartmental clearances Q2 and Q3, whereas Table 3 and the Results text
  label them Q3 and Q4. Table 3 and the Results agree with each other and with
  the volume labels V2/V3/V4, so that numbering is used; the mapping to
  nlmixr2lib names is V2 -> `vc`, V3 -> `vp` with Q3 -> `q`, and V4 -> `vp2`
  with Q4 -> `q2`. Note that `peripheral1` is the large slowly-equilibrating
  compartment (66.0 L at 2.6 L/h) and `peripheral2` the small fast one (10.7 L
  at 5.2 L/h); this follows the paper's numbering rather than a fast-then-slow
  convention.
* **Screened but unretained covariates.** Age, sex, race, ethnicity, migraine
  status, creatinine clearance, oral contraceptives and sumatriptan were tested
  and not retained (Results "Covariate analysis"). Those with an existing
  canonical column are documented in the model's `covariatesDataExcluded`
  metadata; they are deliberately absent from `covariateData` because the paper
  publishes no point estimate for any of them.
* **Supplement not on disk.** Comisar 2025 cites Table S1 (bioanalytical method
  validation), Table S2 (covariates tested), Figure S1 (observed concentrations)
  and Figure S2 (external-validation VPC). None of these is on disk. None is a
  parameter source: Table 3 is complete, and every derived quantity quoted in
  the Discussion reproduces from it, so the extraction does not depend on the
  supplement.
* **Virtual cohort.** Original observed data are not available. Body weight is
  drawn log-normally with median 74 kg truncated to the published 49-131 kg
  range; all other covariates are set to the arm definition rather than sampled,
  because the validation targets above are covariate-specific contrasts rather
  than population VPCs.
* **No non-paper-derived parameter values.** Every `ini()` value comes from
  Comisar 2025 Table 3 or from the Methods statement of the fixed allometric
  exponents. No value was digitised from a figure, supplied by correspondence,
  or carried from an upstream model.
