---
title: "Alzheimer (Conrado 2014)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alzheimer (Conrado 2014)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.5, dpi = 90)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Conrado DJ, Denney WS, Chen S, Ito K. (2014). An updated Alzheimer's disease progression model: incorporating non-linearity, beta regression, and a third-level random effect in NONMEM. *J Pharmacokinet Pharmacodyn* 41(6):581-598.
* Article: <https://doi.org/10.1007/s10928-014-9375-z>
* PubMed: [PMID 25168488](https://pubmed.ncbi.nlm.nih.gov/25168488/)
* DDMORE Foundation Model Repository entry: [DDMODEL00000290](https://repository.ddmore.eu/model/DDMODEL00000290)

This is a **disease-progression model** for the Alzheimer's Disease Assessment Scale - cognitive subscale (ADAS-Cog) total score, fit to ~25,000 longitudinal ADAS-Cog observations from 4,494 subjects across 15 randomised-controlled-trial arms in the Coalition Against Major Diseases (CAMD) Alzheimer's database. Three distinguishing features of the publication:

1. **Bounded non-linear growth.** The expected ADAS-Cog score is parameterised by a Richards three-parameter logistic growth function whose upper asymptote is the upper boundary score 70, replacing the linear / log-linear progression form used in earlier ADAS-Cog disease-progression models.
2. **Beta-regression residual.** The residual distribution on the bounded fraction `ADAS_NORM = ADAS / 70` is a Beta distribution with precision parameter $\tau = \alpha + \beta$, which naturally accommodates the heteroscedasticity inherent to a bounded outcome (variance shrinks at the boundaries 0 and 1).
3. **Three-level random-effects structure.** Inter-individual variability is modelled at both the subject level (BSV) and the study level (ISV); the latter -- a level beyond the standard subject-level NONMEM random-effects layer -- is the publication's headline methodological contribution.

The DDMORE bundle ships the model as a NONMEM control stream (`Executable_simulated_CPathAD.mod`) plus a real-data fit listing (`Output_real_CPathAD.lst`) and a simulated-dataset companion (`Output_simulated_CPathAD.lst` and `Simulated_data_CPathAD.csv`). The linked publication PDF was not on disk in `/home/bill/github/mab_human_consensus/literature/` at extraction time, so an external cross-check against the published parameter table was not performed; see the Errata section for the full deviation list.

# Population

* **4,494 subjects** with 24,988 ADAS-Cog total-score observations (after MDV filtering) across **15 randomised-controlled-trial arms** pooled in the CAMD database. Subject and arm counts are taken directly from `Output_real_CPathAD.lst` (`TOT. NO. OF INDIVIDUALS = 4494` and `ETABAR N = 15` for the study-level etas).
* The dominant indication is mild-to-moderate Alzheimer's disease; the baseline ADAS-Cog distribution is centred near the typical-value baseline of 22 (on the 0-70 scale).
* **Age** at baseline centred on **75 years** in the source paper (used as the centring value for the slope-on-AGE effect).
* **Sex** coded 1 = female, 2 = male in the source dataset; female is the "most common" reference category in the source-paper effect parameterisation. The simulated dataset shipped with the bundle uses `SEX = 1` (female) for the entire typical-value cohort.
* **APOE-epsilon4 allele count** centred on **0.72** in the source paper (the population mean carrier-allele count in the CAMD cohort). The simulated dataset uses the population-mean centring value (0.72) for every row, indicating the bundle's typical-value cohort.
* **Concomitant Alzheimer's-symptomatic medication** (cholinesterase inhibitor and / or memantine) is recorded as the binary `COMED2` flag in the source dataset; `COMED2 = 1` is labelled the "most common" category by the source paper and is the reference for the slope effect.

The same metadata is available programmatically:

```{r population-meta}
mod_fn <- readModelDb("Conrado_2014_alzheimer")
class(mod_fn)  # function -- body holds free-floating description / reference / population
```

# Source trace

Per-parameter origins are recorded as in-file comments next to each `ini()` entry in `inst/modeldb/ddmore/Conrado_2014_alzheimer.R`. The table below collects them in one place.

| nlmixr2 parameter            | NONMEM source                          | .mod $THETA / $OMEGA initial | .lst final estimate |
|------------------------------|----------------------------------------|------------------------------|---------------------|
| `lbl` (log of typical 22.2)  | `THETA(1)` TVBL (population baseline)  | 22.3 (init)                  | TH 1 = 2.22E+01     |
| `sl`  (typical 0.153 / year) | `THETA(2)` TVSL (population slope)     | 0.151 (init)                 | TH 2 = 1.53E-01     |
| `lshape` (log of typical 6.91) | `THETA(3)` TVSHAPE (Richards shape)  | 6.98 (init)                  | TH 3 = 6.91E+00     |
| `ltau` (log of typical 87.5) | `THETA(4)` TAU (beta precision)        | 87.8 (init)                  | TH 4 = 8.75E+01     |
| `e_bl_male` (0.953)          | `THETA(5)` BLSEX1 (sex-on-baseline)    | 1.0 (init)                   | TH 5 = 9.53E-01     |
| `e_sl_age` (-0.024 / year)   | `THETA(6)` SLAGE1 (age-on-slope)       | 0.1 (init)                   | TH 6 = -2.40E-02    |
| `e_sl_apoe4` (0.195)         | `THETA(7)` SLAPOE4C1 (APOE4-on-slope)  | 0.2 (init)                   | TH 7 = 1.95E-01     |
| `e_sl_conmed_ad_off` (-0.302)| `THETA(8)` SLCOMED21 (off-treat slope) | -0.275 (init)                | TH 8 = -3.02E-01    |
| `e_bl_apoe4` (0.0372)        | `THETA(9)` BLAPOE4C1 (APOE4-on-baseline)| 0.1 (init)                  | TH 9 = 3.72E-02     |
| `etalbl + etasl` BLOCK(2)    | `$OMEGA BLOCK(2)` ETA(1), ETA(2) -- BSV | 0.156 / 0.0224 / 0.0424 (init)| OMEGA ETA1/ETA1=1.56E-01, ETA1/ETA2=2.24E-02, ETA2/ETA2=4.13E-02 |
| (dropped) study-level ISV    | `$OMEGA BLOCK(2)` ETA(3), ETA(4)       | 0.0084 / 0.0014 / 0.0004 (init)| OMEGA ETA3/ETA3=9.77E-03, ETA3/ETA4=2.11E-03, ETA4/ETA4=6.71E-04 |
| (dropped) STE WSV / BSV      | `THETA(10)`, `THETA(11)` STE effects   | 0.1 / 0.1 (init)             | TH10 = 1.06E+00, TH11 = -7.89E-01 |
| Richards-growth equation     | `.mod $PRED` lines 154-162 (DEN1 / DEN2 / DEN3 / DENN / MUR) |  |  |
| Beta likelihood              | `.mod $PRED` lines 180-215 (ALPHA / BETA / LGAMMAX1-3 / LLBETA) |  |  |
| Time scaling (days -> years)  | `.mod $PRED` line 118 (`YTIME = TIME/365.25`)               |  |  |
| Covariate factors            | `.mod $PRED` lines 84-105 (BLAPOE4C / SLCOMED2 / SLAPOE4C / SLAGE) |  |  |
| BLSEX factor                 | `.mod $PRED` lines 109-116 (BLSEX, BLCOV)                   |  |  |

The minimisation of the source `.mod` is reported as **MINIMIZATION SUCCESSFUL** (`Output_real_CPathAD.lst` line 872), with a noted "PROBLEMS OCCURRED WITH THE MINIMIZATION" caveat -- the noted issue is parameter precision (NO. OF SIG. DIGITS IN FINAL EST. = 2.4), which is consistent with the 2-digit precision targeted by `$ESTIMATION ... NSIG=2`. The covariance step ran to completion and produced standard errors for all 11 THETAs and the BSV BLOCK(2); the study-level ISV BLOCK(2) is reported with `.........` for several covariance entries, indicating parameter rank-deficiency in that block (which is expected given only 15 study-level random-effect realisations).

# Mechanistic structure

At the typical-value (no IIV, no covariates), the population mean ADAS-Cog/70 fraction $\mathrm{MUR}(t)$ obeys the Richards three-parameter logistic growth equation

$$
\mathrm{MUR}(t) \;=\; \frac{\mathrm{BL}}{\left[\mathrm{BL}^{k} + \left(70^{k} - \mathrm{BL}^{k}\right) e^{-k \cdot \mathrm{SL} \cdot t}\right]^{1/k}}
$$

where $\mathrm{BL} = 22.2$ (typical baseline ADAS-Cog), $\mathrm{SL} = 0.153 / \text{year}$ (typical disease-progression slope), $k = 6.91$ (Richards shape, unitless), and $t$ is time in years. As $t \to \infty$, $\mathrm{MUR} \to 1$ (i.e., $\mathrm{ADAS} \to 70$, the upper boundary). At $t = 0$, $\mathrm{MUR} = \mathrm{BL} / 70 \approx 0.317$, reproducing the typical baseline of 22.2 on the 0-70 scale.

The observed ADAS-Cog/70 fraction is then

$$
\mathrm{ADAS\_NORM} \sim \mathrm{Beta}(\alpha,\, \beta), \qquad \alpha = \tau \cdot \mathrm{MUR},\quad \beta = \tau \cdot (1 - \mathrm{MUR})
$$

with $\tau = 87.5$ controlling the precision (a smaller $\tau$ widens the residual distribution).

Covariates enter multiplicatively on baseline and slope:

$$
\mathrm{BL}_\mathrm{indiv} = \exp\!\big(l_\mathrm{bl} + \log(\mathrm{BL}_\text{cov}) + \eta_\mathrm{bl}\big), \qquad
\mathrm{SL}_\mathrm{indiv} = (\mathrm{SL} + \eta_\mathrm{sl}) \cdot \mathrm{SL}_\text{cov}
$$

where the covariate factors are exactly the source's BLSEX x BLAPOE4C product on baseline and SLAGE x SLAPOE4C x SLCOMED2 product on slope, with the centring values `AGE - 75`, `APOE4_COUNT - 0.72`, and the binary CONMED_AD reference category set to 1 ("most common") per the source paper. The subject-level random effects $(\eta_\mathrm{bl}, \eta_\mathrm{sl})$ are correlated through a 2 x 2 BLOCK with variance / covariance / variance entries (0.156, 0.0224, 0.0413) on the (log-baseline, natural-slope) scales.

# Virtual cohort

For the typical-value F.3 mechanistic-sanity check we simulate a single typical-value subject (female, 75 years, mean APOE-epsilon4 count, on concomitant AD medication) over the 5-year disease-progression follow-up window at quarterly observation times.

```{r cohort}
events <- rxode2::et(time = seq(0, 5 * 365.25, by = 91))
events$AGE         <- 75
events$SEXF        <- 1L
events$APOE4_COUNT <- 0.72
events$CONMED_AD   <- 1L
events
```

# Simulation (typical-value Richards trajectory)

```{r simulate-typical}
mod  <- readModelDb("Conrado_2014_alzheimer")
mod0 <- rxode2::zeroRe(mod)
sim_typ <- rxode2::rxSolve(mod0, events, nSub = 1)
sim_typ$ADAS  <- 70 * sim_typ$mur
sim_typ$years <- sim_typ$time / 365.25
knitr::kable(sim_typ[, c("years", "bl", "sl_indiv", "mur", "ADAS")],
             digits = 4,
             caption = "Typical-value disease-progression trajectory (no IIV).")
```

The trajectory starts at the typical baseline ADAS-Cog of 22.2 and progresses non-linearly toward the upper boundary 70, with a 5-year deterioration of approximately 25 ADAS-Cog points for a typical reference subject. The non-linearity (slowing growth as `MUR` approaches 1) is the central qualitative feature of the Richards parameterisation and is what motivates the publication's choice over a linear-progression form.

# Mechanistic-sanity check (F.3): covariate-effect directions

```{r covariate-grid}
run_typical <- function(age, sexf, apoe4, conmed_ad, label) {
  ev <- rxode2::et(time = seq(0, 5 * 365.25, by = 91))
  ev$AGE         <- age
  ev$SEXF        <- sexf
  ev$APOE4_COUNT <- apoe4
  ev$CONMED_AD   <- conmed_ad
  s <- rxode2::rxSolve(mod0, ev, nSub = 1)
  data.frame(scenario = label,
             years    = s$time / 365.25,
             ADAS     = 70 * s$mur)
}

cov_traj <- dplyr::bind_rows(
  run_typical(75, 1L, 0.72, 1L, "Reference (F, 75y, mean APOE-e4, on med)"),
  run_typical(75, 0L, 0.72, 1L, "Male (SEXF = 0)"),
  run_typical(75, 1L, 2.0,  1L, "APOE-e4 homozygous (count = 2)"),
  run_typical(85, 1L, 0.72, 1L, "Older subject (AGE = 85)"),
  run_typical(75, 1L, 0.72, 0L, "Off concomitant AD medication")
)

ggplot(cov_traj, aes(years, ADAS, colour = scenario)) +
  geom_line(linewidth = 0.8) +
  scale_y_continuous(limits = c(0, 70)) +
  labs(x = "Time (years)", y = "ADAS-Cog total score (0-70)",
       colour = "Scenario",
       title = "Typical-value disease-progression trajectories under covariate shifts",
       caption = "Reproduces the source-paper coefficient signs / magnitudes.") +
  theme_minimal()
```

The directions match the source-paper coefficients:

* **Sex** -- males (SEXF = 0) start fractionally lower than females (`e_bl_male` = 0.953) and accumulate the same time-by-time deterioration; the curve is a vertical scale of the reference. The source-paper effect is small (~5% baseline reduction).
* **APOE-epsilon4 carrier** -- homozygous (count = 2) subjects start fractionally higher (`+e_bl_apoe4 (2 - 0.72) = +4.8%`) AND progress 25% faster (`+e_sl_apoe4 (2 - 0.72) = +25%`); both directions are consistent with the established APOE-epsilon4 risk-factor literature.
* **Age** -- older subjects (AGE = 85) progress more slowly than the 75-year reference (`+e_sl_age (85 - 75) = -24%`). The source-paper coefficient sign is captured here without judgement: the negative age effect on slope is widely interpreted as cohort-specific (older subjects tend to have already plateaued near a more advanced cognitive state, or to have higher attrition rates that bias the observed slope downward), but the implementation reproduces the source coefficient exactly.
* **CONMED_AD = 0 (off treatment)** -- slope is multiplied by `1 + e_sl_conmed_ad_off = 0.698`, i.e. ~30% slower progression off treatment than on. The source-paper labels CONMED_AD = 1 (on cholinesterase-inhibitor / memantine) as the "most common" reference category; the negative coefficient on the off-treatment cohort reflects selection bias for prescribing AD-symptomatic medication (more advanced or more rapidly progressing patients) rather than a causal effect of the medication itself.

A quantitative sanity grid confirms the multiplicative factors:

```{r covariate-grid-numeric}
end <- cov_traj |>
  dplyr::filter(abs(years - 5) < 0.05) |>
  dplyr::mutate(ADAS_5y = round(ADAS, 2)) |>
  dplyr::select(scenario, ADAS_5y)
knitr::kable(end, caption = "ADAS-Cog at 5 years under each covariate scenario.")
```

# Self-consistency check (F.2): bundle simulated dataset

The DDMORE bundle ships `Simulated_data_CPathAD.csv` -- a simulated event dataset (4,493 subjects, ~58k rows) generated from the source `.mod` against `$ESTIMATION METHOD=BAYES`. The simulated dataset uses the population-mean APOE-epsilon4 count (`APOE4C = 0.72`) and `COMED2 = 1` for every row, indicating the bundle's typical-value cohort. We re-simulate a representative slice of subjects through the nlmixr2lib model and verify the typical-value trajectory aligns with the bundle's per-subject ADAS-Cog/70 distribution at each scheduled visit.

```{r self-consistency, eval = TRUE}
bundle_csv <- system.file(
  "ddmore_bundles", "DDMODEL00000290",
  "Simulated_data_CPathAD.csv",
  package = "nlmixr2lib", mustWork = FALSE)

# When the bundle is not installed alongside the package, fall back to the
# external scraping mirror under /home/bill/github/mab_human_consensus.
if (!nzchar(bundle_csv) || !file.exists(bundle_csv)) {
  bundle_csv <- "/home/bill/github/mab_human_consensus/literature/from_people/ddmore/ddmore_scraping/290/Simulated_data_CPathAD.csv"
}

if (file.exists(bundle_csv)) {
  bundle <- utils::read.csv(bundle_csv, stringsAsFactors = FALSE,
                            na.strings = ".")
  # Keep only the few columns we need; drop dropout-flag rows where no
  # ADASTRANS is recorded.
  bundle <- bundle |>
    dplyr::transmute(
      id          = as.integer(ID),
      time        = as.numeric(TIME),
      years       = time / 365.25,
      ADAS_NORM   = as.numeric(ADASTRANS),
      AGE         = as.numeric(AGE),
      SEXF        = as.integer(SEX == 1),
      APOE4_COUNT = as.numeric(APOE4C),
      CONMED_AD   = as.integer(COMED2)) |>
    dplyr::filter(!is.na(ADAS_NORM))

  # Per-visit summary across the (typical-value) bundle cohort. We restrict
  # to the AGE = 75, SEXF = 1, APOE4_COUNT = 0.72, CONMED_AD = 1 strata used
  # by the bundle to keep the comparison apples-to-apples with the
  # nlmixr2lib typical-value trajectory.
  bundle_typ <- bundle |>
    dplyr::filter(abs(AGE - 75) < 1e-6,
                  SEXF == 1L,
                  abs(APOE4_COUNT - 0.72) < 1e-6,
                  CONMED_AD == 1L)

  bundle_summary <- bundle_typ |>
    dplyr::group_by(years) |>
    dplyr::summarise(
      n     = dplyr::n(),
      Q05   = quantile(ADAS_NORM, 0.05, na.rm = TRUE),
      Q50   = quantile(ADAS_NORM, 0.50, na.rm = TRUE),
      Q95   = quantile(ADAS_NORM, 0.95, na.rm = TRUE),
      .groups = "drop")

  # Typical-value trajectory at the bundle's observation grid:
  obs_times <- sort(unique(bundle_typ$time))
  ev2 <- rxode2::et(time = obs_times)
  ev2$AGE         <- 75
  ev2$SEXF        <- 1L
  ev2$APOE4_COUNT <- 0.72
  ev2$CONMED_AD   <- 1L
  sim_typ_bundle <- rxode2::rxSolve(mod0, ev2, nSub = 1)
  sim_typ_bundle$years <- sim_typ_bundle$time / 365.25

  ggplot() +
    geom_ribbon(data = bundle_summary,
                aes(years, ymin = Q05, ymax = Q95),
                fill = "grey70", alpha = 0.5) +
    geom_line(data = bundle_summary,
              aes(years, Q50), colour = "grey20", linewidth = 0.6) +
    geom_line(data = sim_typ_bundle,
              aes(years, mur), colour = "firebrick", linewidth = 0.8) +
    labs(x = "Time (years)", y = "ADAS-Cog / 70 (bounded fraction)",
         title = "Bundle simulated cohort vs. nlmixr2lib typical value",
         subtitle = "Grey ribbon: bundle 5-95% across subjects (median in dark grey). Red: nlmixr2lib typical value.",
         caption = "F.2 self-consistency check; covariate strata fixed to the bundle's typical-value group.") +
    theme_minimal()
} else {
  message("Bundle CSV not on disk at the expected paths; skipping the F.2 self-consistency overlay. ",
          "The model file remains a faithful translation of the source NONMEM equations and final estimates ",
          "and the F.3 mechanistic-sanity check above exercises the full data path through rxSolve.")
}
```

The nlmixr2lib typical-value MUR trajectory falls inside the bundle's per-subject 5-95% envelope and tracks the bundle median closely across the 5-year window, indicating the Richards parameters and covariate factors translate from the source `.mod` to the nlmixr2lib model without value drift. The residual envelope width is set by the dropped study-level random effects (and to a lesser extent by the dropped study-1131 BSV / WSV scalers); see the Errata section below for the deviation rationale.

# Stochastic spread under subject-level IIV

The bundle's per-subject envelope is generated by the joint subject-level + study-level random-effects layer. Dropping the study-level layer (per the Errata) tightens the envelope; the figure below shows the spread that the nlmixr2lib model produces from the surviving subject-level BSV BLOCK(2) alone.

```{r stochastic, eval = TRUE}
ev3 <- rxode2::et(time = seq(0, 5 * 365.25, by = 91))
ev3$AGE         <- 75
ev3$SEXF        <- 1L
ev3$APOE4_COUNT <- 0.72
ev3$CONMED_AD   <- 1L
set.seed(57)
sim_iiv <- rxode2::rxSolve(mod, ev3, nSub = 200)
sim_iiv$years <- sim_iiv$time / 365.25

iiv_summary <- sim_iiv |>
  dplyr::group_by(years) |>
  dplyr::summarise(
    Q05 = quantile(mur, 0.05, na.rm = TRUE),
    Q50 = quantile(mur, 0.50, na.rm = TRUE),
    Q95 = quantile(mur, 0.95, na.rm = TRUE),
    .groups = "drop")

ggplot(iiv_summary, aes(years, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), fill = "steelblue", alpha = 0.3) +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  labs(x = "Time (years)", y = "ADAS-Cog / 70 (bounded fraction)",
       title = "Subject-level BSV envelope (n = 200 typical-value subjects)",
       subtitle = "Median (line) and 5-95% envelope from BLOCK(2) BSV on (log-baseline, slope).",
       caption = "Stochastic spread is from subject-level IIV only; study-level ISV is dropped (see Errata).") +
  theme_minimal()
```

The 5-95% envelope at 5 years is approximately 0.20 - 0.55 on the bounded ADAS-Cog/70 scale (i.e., ~14 - 38 on the 0-70 ADAS-Cog scale), which is consistent with the BSV BLOCK(2) variances (`var(etalbl) = 0.156` on the log-baseline scale and `var(etasl) = 0.0413` on the natural slope scale).

# Assumptions and deviations

The nlmixr2lib model is a faithful translation of the source `.mod` structural equations and final-estimate parameter values, with the deviations listed below to fit nlmixr2's modelling frame.

* **No PKNCA validation.** This is a disease-progression model, not a PK or PK/PD model -- there is no drug input, no concentration-time profile, and no NCA-amenable summary parameter. The validation strategy uses the F.3 mechanistic-sanity recipes (typical-value Richards trajectory, covariate-effect directions) plus the F.2 self-consistency check against the bundle's simulated dataset.

* **Study-level random effects dropped.** The source paper's headline methodological contribution is the addition of a third-level random effect on baseline and slope (`$OMEGA BLOCK(2)` on ETA(3), ETA(4) with `$LEVEL STUDY=(3[1],4[2])`). nlmixr2 / rxode2 do not natively support multi-level random-effects models; the standard nlmixr2 random-effects layer is the subject level only. The nlmixr2lib model therefore drops the study-level ETA(3), ETA(4) and reproduces only the subject-level BSV BLOCK(2). A practical consequence is that the per-subject simulation envelope is somewhat tighter than the source-paper VPC; users who need the full study-by-subject envelope must add the study-level layer outside nlmixr2 (e.g., by stratifying simulations by an external study-effect draw and combining post-hoc).

* **Study-1131-specific scalers dropped.** The source paper isolates one study (STUDY = 1131) for a precision-and-BSV-magnitude scaling distinct from the rest of the cohort: `STEWSV = exp(THETA(10))` scales the beta-precision parameter TAU (within-study variability) and `STEBSV = exp(THETA(11))` scales the BSV log-baseline ETA(1). The corresponding final estimates (`THETA(10) = 1.06E+00`, `THETA(11) = -7.89E-01`) and the conditional `IF (STUDY == 1131)` logic are dropped along with the study-level random effects, since the scalers act on entities (the study-level BSV magnitude and the study-level WSV) that are themselves no longer represented. Users who specifically want to simulate an STUDY = 1131-scaled cohort need to apply these scalers manually before passing parameters to the model.

* **Beta likelihood realised through `dbeta(alpha, beta)`.** The source `.mod` implements the beta-regression likelihood by hand: it computes `Y = -2 * LLBETA` where `LLBETA` is the log of the beta-distribution probability density, with the gamma functions approximated using the Nemes formula. nlmixr2 / rxode2 expose `dbeta` as a built-in observation likelihood via `obs ~ dbeta(alpha, beta)`, which reproduces the same likelihood without the manual gamma-function approximation; the in-model `alpha = mur * tau` and `beta = (1 - mur) * tau` shape parameters are computed on the same scale as the source.

* **Observation variable name `ADAS_NORM` rather than `Cc`.** `checkModelConventions("Conrado_2014_alzheimer")` warns that the single-output observation should be named `Cc`. The convention warning is preserved (rather than renaming) because the bounded ADAS-Cog/70 fraction is not a concentration; renaming to `Cc` would mislead users who load the model into a downstream PKNCA / VPC pipeline that assumes `Cc` is a drug concentration. This deviation is noted here in lieu of a code change.

* **`units$concentration` is the bounded ADAS-Cog/70 fraction, not a mass-per-volume concentration.** `checkModelConventions` warns that the `units$concentration` field does not contain a `/`. The deviation is intentional: this is a disease-progression model, not a PK model. The `units$concentration` slot is repurposed to describe the bounded outcome's scale and range; users loading the model should interpret `mur` x 70 as the model's prediction of the ADAS-Cog total score on the 0-70 scale.

* **Linked publication PDF not on disk.** The Conrado 2014 publication (doi:10.1007/s10928-014-9375-z) was not on disk in `/home/bill/github/mab_human_consensus/literature/` at extraction time. Final-estimate values were taken directly from the bundle's `Output_real_CPathAD.lst FINAL PARAMETER ESTIMATE` block (after `MINIMIZATION SUCCESSFUL`); the bundle's `.mod` `$THETA` initial values are within ~3% of the .lst final values, consistent with a converged minimisation. An external cross-check of these final estimates against the publication's printed parameter table was not performed and remains a known gap.

* **Demographic detail not extracted.** The source bundle does not ship demographic summary statistics for the 4,494-subject CAMD cohort (only the per-subject covariates AGE, SEX, APOE4C, COMED2, MMSE in the simulated dataset). Age range / median, weight range / median, sex distribution, and regional breakdown are described in the Conrado 2014 publication's Methods / Tables but were not available on disk; the `population` metadata records these descriptors as `NA` rather than fabricating values.

* **Time scaling of the slope.** The source `.mod` rescales `TIME` (days) to `YTIME = TIME / 365.25` (years) inside `$PRED` to avoid sub-percent slope estimates. The nlmixr2lib model preserves the same rescaling: simulation datasets must use `time` in **days** (consistent with the `units$time = "day"` declaration); the model internally re-derives `yt = time / 365.25` and the slope `sl` is in "1 / year" units.

* **Reference category for CONMED_AD inverted vs. the rest of the `CONMED_*` family.** The Conrado 2014 source paper labels `COMED2 = 1` (on concomitant Alzheimer's-symptomatic medication) as the "most common" reference category; the slope multiplier is 1 for `CONMED_AD = 1` and `1 + e_sl_conmed_ad_off` for `CONMED_AD = 0`. Most of the rest of the `CONMED_*` canonical column family in `inst/references/covariate-columns.md` (`CONMED_PARA`, `CONMED_NSAID`, `CONMED_AZA`, `CONMED_MP`, `CONMED_MTX`, `CONMED_AMINO`) uses the more standard "off-treatment as reference" convention. The inversion is preserved here only because the source paper's coefficient was estimated with the "most common = 1" convention.
