---
title: "Magnesium sulfate (da Costa 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Magnesium sulfate (da Costa 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("daCosta_2020_magnesiumSulfate"))$reference`
* Description: `r rxode2::rxode(readModelDb("daCosta_2020_magnesiumSulfate"))$description`
* Article: <https://doi.org/10.1007/s40268-020-00315-2>

da Costa 2020 is a prospective observational population-PK study of
intravenous magnesium sulfate given by the Zuspan regimen (4 g over 30 min,
then 1 g/h) to 109 Brazilian women with preeclampsia in a maternal intensive
care unit. The model is one-compartment with first-order elimination, fitted in
Monolix, with serum creatinine on clearance and body weight on volume. Other
magnesium sulfate models in this library are `Salinger_2013_magnesiumSulfate`,
`Easterling_2018_magnesium_sulfate`, `Du_2019_magnesiumSulfate` and
`Deng_2024_magnesiumSulfate`.

**Read the Assumptions and deviations section before using this model.** The
covariate equation printed in the paper is unusable as written. The centring
used here is supported by the paper's own visual predictive check, as shown
below. The model has no endogenous magnesium baseline, and it
overpredicts the observed magnesium concentrations. The authors say so
themselves.

# Population

```{r population-metadata}
pop <- rxode2::rxode(readModelDb("daCosta_2020_magnesiumSulfate"))$population
str(pop, max.level = 1)
```

Baseline characteristics from da Costa 2020 Table 1 (n = 109, 347 serum
magnesium concentrations, three to four per woman):

| Characteristic | Mean +/- SD |
|---|---|
| Age (years) | 25.8 +/- 7.4 |
| Body weight (kg) | 79.2 +/- 14.7 |
| Baseline magnesium (mg/dL) | 1.9 +/- 0.6 |
| Gestational age (weeks) | 35.2 +/- 4.2 |
| Baseline creatinine (mg/dL) | 0.7 +/- 0.3 |

# Source trace

| Element | Value in model | Source |
|---|---|---|
| Structure: one compartment, first-order elimination, IV infusion | `d/dt(central) <- -kel * central` | Methods 2.4 (infusion equation); Table 2 (1-cmt combined error selected) |
| `lcl` | log(1.38) L/h | Table 3, CL_pop; Results equation `exp(0.3221 ...)` |
| `lvc` | log(13.3) L | Table 3, V_pop; Results equation `exp(2.5878 ...)` |
| `e_creat_cl` | -0.0814 per mg/dL | Table 3, Beta_CL_creatinine_mg_dL |
| `e_wt_vc` | 0.0752 per kg | Table 3, Beta_V_Weith_kg |
| `etalcl` | 0.015^2 = 0.000225 | Table 3, Omega_CL = 0.015 (Monolix SD) |
| `etalvc` | 0.404^2 = 0.163216 | Table 3, Omega_V = 0.404 (Monolix SD) |
| `addSd` | 1.76 mg/dL | Table 3, residual error a |
| `propSd` | 0.000552 | Table 3, residual error b |
| Covariate centring WT - 79.2, CREAT - 0.7 | cohort means | Table 1; supported by Figure 3 (see below) |
| Concentration unit mg/dL, no baseline | `Cc <- central / vc / 10` | Figure 3 axis; Methods 2.4 'No endogenous magnesium baseline adjustment'; Figure 1 predictions of 0 at pre-dose samples |

# Units

Doses are entered as **mg of elemental magnesium**, matching the other
magnesium models in this library: multiply grams of MgSO4-7H2O by
24.305 / 246.47 = 0.0986. The 4 g loading dose is 394.4 mg Mg, and the 1 g/h
maintenance infusion is 98.6 mg Mg/h. `Cc` is in mg/dL of elemental
magnesium, the unit of the paper's VPC axis and of the residual error `a`.

```{r units}
mg_per_g <- 24.305 / 246.47 * 1000
load_mg <- 4 * mg_per_g
rate_mg <- 1 * mg_per_g
c(load_mg = load_mg, rate_mg_per_h = rate_mg)
```

# Covariate centring: the printed equation versus Figure 3

The Results print
`V = exp(2.5878 + 0.0752 x weight (kg) + 0.404)`. Taken literally, this adds
the omega to the exponent and uses weight uncentred. At the cohort-mean weight
it gives V = 13.3 x exp(0.0752 x 79.2) = about 5,000 L. Magnesium would then
be undetectable, so that reading is not tenable. The Monolix beta must act on
a centred covariate. The paper does not print the centring value.

Figure 3, a VPC that simulates from the final model, settles the question. Its
median prediction line starts at 0 at time 0 and rises through 1.3 mg/dL at
2 h and 3.4 mg/dL at 6 h. These medians were digitised from the figure by the
maintainers. The rise is reproduced by the 1 g/h infusion alone, which also
shows that the VPC dosing carried no loading dose. With that dosing, only
centring at the cohort mean reproduces it. Centring at a rounded 70 kg makes
the typical V twice as large and halves the early rise.

```{r centring}
mod <- rxode2::rxode(readModelDb("daCosta_2020_magnesiumSulfate"))
mod_typ <- rxode2::zeroRe(mod)
stopifnot(is.null(mod$linCmt))

vpc_median <- data.frame(
  time = c(0, 2, 6, 12, 18, 28),
  fig3_median = c(0, 1.32, 3.43, 5.45, 3.75, 2.11)
)

typ_infusion <- function(wt, creat = 0.7, loading = FALSE) {
  ev <- rxode2::et(amt = rate_mg * 12, rate = rate_mg, cmt = "central")
  if (loading) {
    ev <- rxode2::et(amt = load_mg, dur = 0.5, cmt = "central") |>
      rxode2::et(amt = rate_mg * 12, rate = rate_mg, time = 0.5, cmt = "central")
  }
  ev <- rxode2::et(ev, vpc_median$time, cmt = "central")
  s <- rxode2::rxSolve(mod_typ, ev, params = data.frame(WT = wt, CREAT = creat))
  as.data.frame(s)$Cc
}

# Centring at 70 kg is equivalent, for the typical woman at 79.2 kg, to
# evaluating the 79.2-centred model at 79.2 + (79.2 - 70) kg.
centring <- vpc_median |>
  dplyr::mutate(
    `centred at 79.2 kg` = typ_infusion(79.2),
    `centred at 70 kg` = typ_infusion(79.2 + (79.2 - 70)),
    `79.2 kg, with 4 g load` = typ_infusion(79.2, loading = TRUE)
  )
knitr::kable(centring, digits = 2,
             caption = "Typical-value Cc (mg/dL) under a 1 g/h infusion stopped at 12 h, against the Figure 3 median.")

rise <- centring |> dplyr::filter(time %in% c(2, 6))
stopifnot(
  # Centring at the cohort mean reproduces the published rise within 5%.
  all(abs(rise$`centred at 79.2 kg` / rise$fig3_median - 1) < 0.05),
  # A 70 kg centring misses it by more than 40%, as does a dosing record with
  # the 4 g load.
  all(abs(rise$`centred at 70 kg` / rise$fig3_median - 1) > 0.40),
  abs(rise$`79.2 kg, with 4 g load`[1] / rise$fig3_median[1] - 1) > 0.40
)
```

After 12 h the Figure 3 median declines more slowly than the typical-value
solve (3.75 against 2.73 mg/dL at 18 h). The paper says the VPCs were
stratified by infusion duration, so the pooled median after 12 h mixes women
whose infusion was still running. The dosing history behind that part of the
figure is not reported, so the decline is not used as a check.

# Virtual cohort and Figure 3 replication

The cohort draws weight from N(79.2, 14.7) kg, truncated to 50-120 kg by
rejection (the study excluded BMI > 40 kg/m^2). Creatinine is drawn
log-normally with mean 0.7 and SD 0.3 mg/dL. Every woman receives the 1 g/h
infusion without the loading dose, stopped at 12 h, to match the VPC dosing
inferred above.

```{r cohort}
rxode2::rxSetSeed(20200708)
n_sub <- 200
draw_trunc <- function(n, mean, sd, lo, hi) {
  out <- numeric(0)
  while (length(out) < n) {
    x <- rnorm(n, mean, sd)
    out <- c(out, x[x >= lo & x <= hi])
  }
  out[seq_len(n)]
}
creat_sdlog <- sqrt(log(1 + (0.3 / 0.7)^2))
cohort <- data.frame(
  id = seq_len(n_sub),
  WT = draw_trunc(n_sub, 79.2, 14.7, 50, 120),
  CREAT = rlnorm(n_sub, log(0.7) - creat_sdlog^2 / 2, creat_sdlog)
)

obs_times <- sort(unique(c(seq(0, 28, by = 0.5), vpc_median$time)))
ev_vpc <- rxode2::et(amt = rate_mg * 12, rate = rate_mg, cmt = "central") |>
  rxode2::et(obs_times, cmt = "central") |>
  as.data.frame()
ev_vpc <- merge(cohort["id"], ev_vpc[, setdiff(names(ev_vpc), "id")], by = NULL) |>
  dplyr::left_join(cohort, by = "id") |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim_vpc <- rxode2::rxSolve(mod, ev_vpc, keep = c("WT", "CREAT")) |>
  as.data.frame()

vpc_sum <- sim_vpc |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    p05 = quantile(sim, 0.05), p50 = quantile(sim, 0.50), p95 = quantile(sim, 0.95),
    .groups = "drop"
  )

ggplot(vpc_sum, aes(time)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = p50)) +
  geom_point(data = vpc_median, aes(time, fig3_median), colour = "firebrick", size = 2) +
  labs(x = "Time (h)", y = "Magnesium (mg/dL)",
       title = "Replicates Figure 3 of da Costa 2020",
       caption = "Line and ribbon: simulated median and 5th-95th centiles (with residual error).\nRed points: Figure 3 median prediction line, digitised by the maintainers.")
```

```{r vpc-check}
vpc_cmp <- sim_vpc |>
  dplyr::filter(time %in% c(2, 6)) |>
  dplyr::group_by(time) |>
  dplyr::summarise(ipred_median = median(ipredSim), .groups = "drop") |>
  dplyr::inner_join(vpc_median, by = "time")
knitr::kable(vpc_cmp, digits = 2)
# Envelope check on the cohort's individual-prediction median. The residual
# error (additive SD 1.76 mg/dL) is larger than the 2 h median itself, so the
# median of the noisy simulated data is too unstable to gate on. The weight
# effect alone spreads V by a log-SD of about 1.1, so even the IPRED median of
# 200 women moves by about 10% between cohorts; the bound allows for that. The
# typical-value check above is the tight one.
stopifnot(all(abs(vpc_cmp$ipred_median / vpc_cmp$fig3_median - 1) < 0.35))
```

The simulated 5th-95th band at time 0 (about -3 to +3 mg/dL) is wider than
the Figure 3 band (about -2.2 to +2.3 mg/dL). At time 0 only the additive
error acts, so the published band implies an additive SD near 1.35 mg/dL
against the tabulated a = 1.76. The Table 3 value is kept.

The weight effect is steep. exp(0.0752) = 1.078, so V rises 7.8% per kg, and
one cohort SD of weight (14.7 kg) triples V. It is the dominant source of
between-woman variability in the model, ahead of the 40% log-SD on V.

# The clinical Zuspan regimen

A typical woman receives the protocol regimen (4 g over 30 min, then 1 g/h to
18 h). The paper reports an observed mean steady-state serum magnesium of
1.31 mmol/L (3.18 mg/dL, Discussion). That is total magnesium, and the
baseline is 1.9 mg/dL (Table 1), so the drug-derived part is about 1.3 mg/dL.

```{r zuspan}
ev_z <- rxode2::et(amt = load_mg, dur = 0.5, cmt = "central") |>
  rxode2::et(amt = rate_mg * 17.5, rate = rate_mg, time = 0.5, cmt = "central") |>
  rxode2::et(seq(0, 18, by = 0.25), cmt = "central")
sim_z <- rxode2::rxSolve(mod_typ, ev_z, params = data.frame(WT = 79.2, CREAT = 0.7)) |>
  as.data.frame()
css_typ <- rate_mg / exp(log(1.38)) / 10
knitr::kable(
  sim_z |> dplyr::filter(time %in% c(2, 6, 12, 18)) |> dplyr::select(time, Cc),
  digits = 2,
  caption = "Typical woman (79.2 kg, creatinine 0.7 mg/dL), drug-derived Cc (mg/dL)."
)
c(model_css_mg_dL = css_typ, observed_drug_derived_mg_dL = 3.18 - 1.9)
```

The model's steady state for a typical woman is `r round(css_typ, 1)` mg/dL
of drug-derived magnesium. That is more than five times the observed drug-derived
level. The overprediction is visible in the paper itself: in Figure 3 the
observed concentrations at 12 h sit mostly below the median prediction line,
and the Discussion notes that 'the individual predicted concentrations are
skewed towards being greater than the observed concentrations'. The packaged
model reproduces the published parameters and does not correct this. Do not
use it to predict absolute serum magnesium without that caveat.

# PKNCA validation

NCA of the 4 g loading dose alone, typical values, at three weights. The model
is linear and one-compartment, so the NCA clearance, Dose / AUCinf, must
return the model's CL, and the half-life must equal ln(2) V / CL. `Cc` is in
mg/dL, so CL in L/h is Dose / (10 x AUCinf).

```{r pknca}
wts <- c(60, 79.2, 100)
# Sample each weight for 12 typical half-lives. A fixed long grid drives the
# lightest woman (t1/2 about 1.6 h) far below the solver's absolute tolerance,
# and those noise-level points bias the lambda-z fit.
hl_typ <- log(2) * 13.3 * exp(0.0752 * (wts - 79.2)) / 1.38
ev_sd <- dplyr::bind_rows(lapply(seq_along(wts), function(i) {
  obs <- c(seq(0, 1, by = 0.05), seq(1.25, 12 * hl_typ[i], length.out = 400))
  rxode2::et(amt = load_mg, dur = 0.5, cmt = "central") |>
    rxode2::et(obs, cmt = "central") |>
    as.data.frame() |>
    dplyr::mutate(id = i)
})) |>
  dplyr::mutate(WT = wts[id], CREAT = 0.7, treatment = paste0("WT ", wts[id], " kg")) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim_sd <- rxode2::rxSolve(mod_typ, ev_sd, keep = c("treatment", "WT")) |>
  as.data.frame()

conc_nca <- sim_sd |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)
stopifnot(all(tapply(conc_nca$time, conc_nca$id, min) == 0))
dose_nca <- ev_sd |>
  dplyr::filter(evid != 0) |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(conc_nca, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_nca, amt ~ time | treatment + id)
intervals <- data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                        aucinf.obs = TRUE, half.life = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(treatment, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::inner_join(sim_sd |> dplyr::distinct(id, cl, vc), by = "id") |>
  dplyr::mutate(
    cl_nca = load_mg / (10 * aucinf.obs),
    hl_model = log(2) * vc / cl
  )

nca_wide |>
  dplyr::select(treatment, cmax, tmax, aucinf.obs, half.life, hl_model, cl_nca, cl) |>
  dplyr::rename("Group" = treatment, "Cmax (mg/dL)" = cmax, "Tmax (h)" = tmax,
                "AUCinf (mg*h/dL)" = aucinf.obs, "t1/2 NCA (h)" = half.life,
                "t1/2 model (h)" = hl_model, "CL NCA (L/h)" = cl_nca,
                "CL model (L/h)" = cl) |>
  knitr::kable(digits = 3, caption = "PKNCA on the typical-value 4 g loading dose.")

stopifnot(
  all(abs(nca_wide$cl_nca / nca_wide$cl - 1) < 0.01),
  all(abs(nca_wide$half.life / nca_wide$hl_model - 1) < 0.01)
)
```

The paper reports no NCA parameters (no Cmax, AUC or numeric half-life), so
there is no published NCA table to compare against. The typical half-life at
the cohort-mean weight is ln(2) x 13.3 / 1.38 = 6.7 h.

# Assumptions and deviations

* **Covariate centring (not printed).** The Results equations print weight and
  creatinine uncentred, which makes V about 5,000 L at the cohort-mean weight.
  The model centres both on the Table 1 cohort means (79.2 kg, 0.7 mg/dL).
  The early rise of the Figure 3 VPC supports the weight centring: the
  typical-value solve centred at 79.2 kg matches the published median at 2 and
  6 h within 5%, and a 70 kg centring misses it by more than 40%. The
  creatinine centring cannot be tested this way. Its effect is small
  (exp(-0.0814 x 0.7) = 0.945), and it is centred like weight for consistency.
* **Omega values added into the exponent.** The Results equations add 0.0151
  (CL) and 0.404 (V) inside `exp()`. These are the Table 3 omegas. They are
  treated as the random-effect SDs they are, not as fixed shifts. The Table 3
  value 0.015 is used for omega_CL; the equation's 0.0151 agrees to rounding.
* **Omega scale.** Monolix reports omega as the SD of the log-normal random
  effect, so the variances are 0.404^2 and 0.015^2. The Table 3 'CV' column
  (19.3, 38.5) is the relative standard error of those estimates, not a
  between-subject CV. For omega_V, 0.0777 / 0.404 = 19.2%.
* **Residual error.** Monolix's combined model, sd = a + b f, is encoded as
  nlmixr2 `combined1()`. With b = 0.000552 it is effectively additive at
  1.76 mg/dL. The printed concentration equation,
  `Mg = basalMg + (1.76 + 0.000552 x basalMg) x exp(CL/V x t)`, puts the
  residual-error parameters in place of an amplitude. It is not a usable
  structural equation, and the Methods infusion equation is used instead.
* **Units.** Dose is mg of elemental Mg and `Cc` is mg/dL. The paper does not
  state its dataset's dose unit. The Figure 3 check above fixes the scale: a
  98.6 mg Mg/h infusion reproduces the published rise with the Table 3 CL and
  V. A dose in grams of MgSO4-7H2O would change it by a factor of about 10.
* **No baseline.** Following the paper ('No endogenous magnesium baseline
  adjustment was made to the model'), `Cc` is drug-derived magnesium only.
  The fitted observations were total serum magnesium (baseline
  1.9 +/- 0.6 mg/dL). The abstract's 'baseline magnesium concentration of
  0.77 mmol/L (1.87 mg/dL)' is not a parameter in Table 3.
* **Loading dose in the VPC.** The Figure 3 median starts at 0 and has the
  shape of a pure 1 g/h infusion. With the 4 g load included, the 2 h value
  would be 3.5 mg/dL instead of 1.3 mg/dL. The paper does not say why the load
  is absent. The model file places no restriction on dosing.
* **Overprediction.** As shown above, the model's steady state is more than five
  times the observed drug-derived magnesium. This is a property of the
  published estimates, which the authors acknowledge in their limitations.
* **Inconsistent RSE column.** The Table 3 '%' column for V_pop and CL_pop
  (8.43, 17.4) is ten times the ratio of the printed SE to the estimate
  (0.84%, 1.7%). This does not affect the point estimates.
* No erratum was found in a EuropePMC search on 2026-09-27.
