---
title: "IPSS in benign prostatic hyperplasia: tamsulosin, dutasteride and combination (D'Agate 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IPSS in benign prostatic hyperplasia: tamsulosin, dutasteride and combination (D'Agate 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
ui <- rxode2::rxode(readModelDb("DAgate_2020_bph_ipss_mbma"))
```

* Citation: `r ui$reference`
* Article: <https://doi.org/10.1111/bcp.14268>
* Supplement (Supporting Information, including the final NONMEM control
  stream as an appendix): <https://doi.org/10.1111/bcp.14268>

D'Agate 2020 is an individual-patient-data model-based meta-analysis of six
phase III/IV dutasteride trials (ARIA3001, ARIA3002, ARI40002, CombAT, CONDUCT,
ARIB3003; 10,238 men, 140,733 IPSS observations, up to 4 years). It describes the
time course of the International Prostate Symptom Score (IPSS, 0-35) with one
ordinary differential equation (Equations 1-3):

$$\frac{dIPSS}{dt} = DISP\left(1 - \frac{IPSS}{35}\right) - (PLACEBO + TREATMENT)\cdot IPSS,\qquad IPSS(0) = IPSS_b$$

$$PLACEBO = \Delta_{placebo}\, e^{-t \ln 2 / T_{1/2}},\qquad TREATMENT = \Delta_{treatment}$$

The first term is a zero-order worsening of symptoms (the disease progression
rate `DISP`) that shuts off as the score approaches the scale maximum of 35. The
second term is a first-order improvement: a placebo effect that wears off with a
7.3-month half-life, plus a constant effect for the active treatment (tamsulosin,
dutasteride, watchful waiting, or tamsulosin-dutasteride combination therapy).

# Population

```{r population}
str(ui$population)
```

Table 1 of the paper: men aged 47-94 years (median 66), body weight median 82 kg,
baseline IPSS median 16 (range 1-35; 66% moderate, 29% severe at the start of
treatment), median prostate volume 48.6 cm^3, median PSA 3.4 ng/mL, median BPH
symptom duration 4 years; predominantly white (9268 of 10,147 with race recorded).
Placebo was given to 2158 men for 2 years; 373 were in the CONDUCT
watchful-waiting arm, 1611 received tamsulosin 0.4 mg, 3790 dutasteride 0.5 mg
and 2143 the combination, for up to 4 years.

# Source trace

The control stream in the Supporting Information runs in days; Tables 2 and 3
report the same quantities per month. The conversion factor is 30 days/month:
`THETA(1) = 0.0116 /day` x 30 = 0.348 (Table 2: 0.347), `THETA(2) = 0.00202 /day`
x 30 = 0.0606 (Table 2: 0.061), `THETA(3) = 218 days` / 30 = 7.27 (Table 2:
7.263). The model uses the Table 2 / Table 3 values directly, in months.

| Quantity | Model name | Value | Source |
|---|---|---|---|
| Structural ODE | `d/dt(ipss)` | Eq. 1 | Eq. 1; control stream `DADT(1)` |
| Placebo effect | `eff_pbo` | Eq. 2 | Eq. 2; control stream `EFFpl` |
| Treatment effect | `eff_trt` | Eq. 3 | Eq. 3; control stream `EFFtrt` by `ARM` |
| Initial condition | `ipss(0) <- IPSS_BL` | observed baseline | control stream `A_0(1) = obsIPSS` |
| DISP | `ldisp` | 0.347 points/month | Table 2 |
| DELTA placebo | `ldelta_pbo` | 0.061 /month | Table 2 |
| Placebo T1/2 | `lthalf_pbo` | 7.263 months | Table 2 |
| BPHDUR on placebo magnitude | `e_t_sympt_bph_delta_pbo` | -0.025 /year, centred at 4 y | Table 2; linear form and centre from control stream `DELTpBPHDUR` |
| IPSSb on DISP | `e_ipss_bl_disp` | 0.027 /point, centred at 16 | Table 2; linear form and centre from control stream `DISPIPSS0` |
| Alcohol use on placebo T1/2 | `e_alcohol_use_thalf_pbo` | -0.135 | Table 2; form from control stream `T12pALCU` |
| DELTA tamsulosin | `ldelta_tamsulosin` | 0.015 /month | Table 3 |
| DELTA dutasteride | `ldelta_dutasteride` | 0.016 /month | Table 3 |
| DELTA watchful waiting | `ldelta_ww` | 0.018 /month | Table 3 |
| DELTA combination | `ldelta_fdc` | 0.032 /month | Table 3 |
| IIV DISP, placebo, T1/2 (variances) | `etaldisp`, `etaldelta_pbo`, `etalthalf_pbo` | 0.997, 2.822, 2.758 | Table 2 |
| Correlations in the placebo block | off-diagonals | -0.544, 0.307, -0.965 | control stream `$OMEGA BLOCK(3)` |
| IIV treatment effects (variances) | `etaldelta_*` | tamsulosin 1.923, dutasteride 1.734, WW 1.812, combination 1.449 | Table 3 |
| Additive residual SD | `addSd_*` | placebo 3.224, tamsulosin 3.613, dutasteride 3.506, WW 2.778, combination 3.141 | Tables 2 and 3 |

```{r params}
ini_df <- as.data.frame(ui$iniDf)
ini_df |>
  dplyr::filter(is.na(neta1)) |>
  dplyr::transmute(
    Parameter = name,
    Estimate = signif(ifelse(grepl("^l", name), exp(est), est), 4),
    Scale = ifelse(grepl("^l", name), "back-transformed", "as estimated"),
    Fixed = fix
  ) |>
  knitr::kable(caption = "Fixed effects and residual SDs (log-scale parameters back-transformed).")
```

# Structural checks

## Unit conversion between the control stream and the tables

```{r unit-check}
unit_chk <- data.frame(
  quantity = c("DISP", "DELTA placebo", "placebo T1/2"),
  control_stream_per_day = c(0.0116, 0.00202, 218),
  converted_per_month = c(0.0116 * 30, 0.00202 * 30, 218 / 30),
  table2 = c(0.347, 0.061, 7.263)
)
unit_chk$pct_diff <- 100 * (unit_chk$converted_per_month / unit_chk$table2 - 1)
knitr::kable(unit_chk, digits = 4)
stopifnot(all(abs(unit_chk$pct_diff) < 1))
```

## Long-term steady state is closed-form

Once the placebo effect has decayed, a constant treatment effect `E` gives
`DISP (1 - I/35) = E I`, so the score settles at `I* = DISP / (E + DISP/35)`.
Solving the typical-value model for 100 years must land on that value.

```{r steady-state}
mod_typ <- rxode2::zeroRe(ui)

arm_cov <- function(arm) {
  data.frame(
    TRT_TAMSULOSIN = as.integer(arm == "Tamsulosin"),
    TRT_DUTASTERIDE = as.integer(arm == "Dutasteride"),
    TRT_TAMSULOSIN_DUTASTERIDE = as.integer(arm == "Combination"),
    TRT_WATCHFUL_WAITING = as.integer(arm == "Watchful waiting")
  )
}
arms <- c("Placebo", "Tamsulosin", "Dutasteride", "Combination", "Watchful waiting")

ss_ev <- do.call(rbind, lapply(seq_along(arms), function(i) {
  cbind(
    data.frame(id = i, time = c(0, 1200), evid = 0, cmt = "ipss",
               IPSS_BL = 16, T_SYMPT_BPH = 4, ALCOHOL_USE = 0),
    arm_cov(arms[i])
  )
}))
ss_sim <- as.data.frame(rxode2::rxSolve(mod_typ, ss_ev, returnType = "data.frame"))

disp_typ <- 0.347
eff <- c(Placebo = 0, Tamsulosin = 0.015, Dutasteride = 0.016,
         Combination = 0.032, `Watchful waiting` = 0.018)
ss_chk <- ss_sim |>
  dplyr::filter(time == 1200) |>
  dplyr::mutate(arm = arms[id], closed_form = disp_typ / (eff[arm] + disp_typ / 35)) |>
  dplyr::select(arm, simulated = ipss, closed_form)
knitr::kable(ss_chk, digits = 3)
# Same parameters on both sides: pure numerical error, so a tight bound is right.
stopifnot(all(abs(ss_chk$simulated - ss_chk$closed_form) < 1e-3))
```

Untreated (placebo) patients converge on the scale maximum of 35; the
combination holds a typical patient at about 8 points indefinitely.

# Virtual cohort

Covariate distributions follow Table 1: baseline IPSS normal with mean 16.48 and
SD 6.1 truncated to 1-35; BPH symptom duration log-normal with median 4 years and
mean 5.17 years (so log-SD 0.71), truncated at the observed maximum of 54.8 years;
alcohol use in 6198 of 10,190 men (60.8%).

```{r cohort}
set.seed(2020)
rxode2::rxSetSeed(2020)
n_per_arm <- 200
make_cohort <- function(n, arm, id0) {
  cbind(
    data.frame(
      id = id0 + seq_len(n),
      arm = arm,
      IPSS_BL = pmin(pmax(round(rnorm(n, 16.48, 6.1)), 1), 35),
      T_SYMPT_BPH = pmin(rlnorm(n, log(4), 0.71), 54.8),
      ALCOHOL_USE = rbinom(n, 1, 6198 / 10190)
    ),
    arm_cov(arm)[rep(1, n), , drop = FALSE]
  )
}
cohort <- do.call(rbind, lapply(seq_along(arms), function(i) {
  make_cohort(n_per_arm, arms[i], (i - 1) * n_per_arm)
}))
rownames(cohort) <- NULL
cohort |>
  dplyr::group_by(arm) |>
  dplyr::summarise(n = dplyr::n(), IPSS_BL_median = median(IPSS_BL),
                   T_SYMPT_BPH_median = median(T_SYMPT_BPH),
                   alcohol_pct = 100 * mean(ALCOHOL_USE)) |>
  knitr::kable(digits = 1)
```

# Simulation

Observation visits follow the trials' IPSS schedule (months 0, 1, 3, 6, then
every 3 months). Placebo is simulated for 24 months, as in the source trials, and
the active arms for 48 months.

```{r simulate}
obs_times <- c(0, 1, 3, seq(6, 48, by = 3))
ev <- cohort |>
  tidyr::crossing(time = obs_times) |>
  dplyr::filter(!(arm == "Placebo" & time > 24)) |>
  dplyr::mutate(evid = 0, cmt = "ipss") |>
  dplyr::arrange(id, time)

sim <- as.data.frame(rxode2::rxSolve(ui, ev, returnType = "data.frame",
                                     keep = c("arm")))
# The authors truncated simulated observations to the 0-35 scale
# (control stream $ERROR, ICALL = 4).
sim$sim_obs <- pmin(pmax(sim$sim, 0), 35)
```

```{r vpc, fig.height = 6}
# Replicates the structure of Figure 3 (left panels) of D'Agate 2020:
# simulated median and 95% prediction interval of IPSS by treatment.
vpc <- sim |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(
    p025 = quantile(sim_obs, 0.025), p50 = median(sim_obs),
    p975 = quantile(sim_obs, 0.975), ipred50 = median(ipss), .groups = "drop"
  ) |>
  dplyr::mutate(arm = factor(arm, levels = arms))
ggplot(vpc, aes(time, p50)) +
  geom_ribbon(aes(ymin = p025, ymax = p975), alpha = 0.2) +
  geom_line() +
  geom_line(aes(y = ipred50), linetype = "dashed") +
  facet_wrap(~arm) +
  labs(x = "Time since start of treatment (months)", y = "IPSS",
       title = "Simulated IPSS by treatment arm (200 patients per arm)",
       caption = "Solid: median simulated observation; dashed: median individual prediction; band: 95% PI.") +
  theme_bw()
```

# Replicate published figures

## Typical-patient trajectories (Figures 5-7)

```{r typical-grid-build}
grid <- tidyr::crossing(IPSS_BL = c(8, 12, 16, 20, 30), arm = arms)
grid$id <- seq_len(nrow(grid))
typ_ev <- do.call(rbind, lapply(seq_len(nrow(grid)), function(i) {
  cbind(
    data.frame(id = grid$id[i], time = 0:48, evid = 0, cmt = "ipss",
               IPSS_BL = grid$IPSS_BL[i], T_SYMPT_BPH = 4, ALCOHOL_USE = 0),
    arm_cov(grid$arm[i])
  )
}))
typ <- as.data.frame(rxode2::rxSolve(mod_typ, typ_ev, returnType = "data.frame"))
typ <- dplyr::select(typ, -dplyr::any_of("IPSS_BL")) |> dplyr::left_join(grid, by = "id")

ggplot(dplyr::filter(typ, arm %in% c("Placebo", "Tamsulosin", "Combination")),
       aes(time, ipss, colour = arm)) +
  geom_line() +
  facet_wrap(~IPSS_BL, labeller = label_both, nrow = 1) +
  labs(x = "Time (months)", y = "IPSS (typical patient)",
       title = "Typical IPSS trajectories by baseline severity (cf. Figures 6 and 7)",
       colour = NULL) +
  theme_bw() + theme(legend.position = "bottom")
```

## IPSS difference at 48 months versus untreated (Tables S2-S5)

Supporting Information Tables S3 (combination) and S5 (tamsulosin, 50th
percentile of DISP) report the median difference at 48 months between the
treated and the untreated trajectory. The Table S4 / S5 "moderate" and "severe"
columns reproduce the Table S3 IPSSb = 12 and 20 rows exactly (-8.42 and
-11.38), so they are taken as baseline IPSS 12 and 20.

```{r table-s3}
d48 <- typ |>
  dplyr::filter(time == 48) |>
  dplyr::select(IPSS_BL, arm, ipss) |>
  tidyr::pivot_wider(names_from = arm, values_from = ipss) |>
  dplyr::mutate(sim_comb = Combination - Placebo, sim_tam = Tamsulosin - Placebo)

pub <- data.frame(
  IPSS_BL = c(8, 12, 16, 20, 30, 12, 20),
  treatment = c(rep("Combination", 5), "Tamsulosin", "Tamsulosin"),
  source = c(rep("Table S3", 5), "Table S5 moderate", "Table S5 severe"),
  published = c(-7.1, -8.42, -10.21, -11.38, -14.41, -5.22, -7.14)
)
cmp <- pub |>
  dplyr::left_join(d48, by = "IPSS_BL") |>
  dplyr::mutate(simulated = ifelse(treatment == "Combination", sim_comb, sim_tam),
                pct_diff = 100 * (simulated / published - 1)) |>
  dplyr::select(source, treatment, IPSS_BL, published, simulated, pct_diff)
cmp |>
  dplyr::rename("Source" = source, "Treatment" = treatment, "Baseline IPSS" = IPSS_BL,
                "Published median" = published, "Typical-value model" = simulated,
                "% difference" = pct_diff) |>
  knitr::kable(digits = 2)
stopifnot(all(abs(cmp$pct_diff) < 15), all(cmp$simulated < 0))
```

The typical-value model reproduces every published median to within 4-10%,
consistently slightly smaller in magnitude. The published values are medians of
200 stochastic simulations whose random-effect settings are not stated (Figure 5
fixes DISP at chosen percentiles and draws the other parameters), whereas the
table above uses typical values throughout with a 4-year symptom duration and no
alcohol use, so an exact match is not expected. The ordering by baseline
severity, the roughly two-fold ratio of combination to tamsulosin benefit, and
the magnitudes all agree.

# PKNCA validation

Not applicable. The model has no drug concentrations: treatments enter as arm
indicators with constant effects, because the source trials collected no
pharmacokinetic data (Discussion 4.1). The structural checks above (closed-form
steady state and unit conversion) and the Table S2-S5 comparison replace the NCA
comparison.

# Assumptions and deviations

* **Placebo random-effect block.** Table 2 reports only the variances of the
  three placebo-model random effects (0.997, 2.822, 2.758). The Supporting
  Information states that the final model carried eta correlations between DISP
  and the placebo magnitude and half-life, and its control stream fixes an
  `$OMEGA BLOCK(3)` of `0.982; -0.977, 3.29; 0.524, -3.01, 2.96`, whose diagonal
  differs slightly from Table 2 (it was most likely carried over from a
  re-estimation on a different data subset). The model keeps the Table 2
  variances and applies the control stream's correlations (-0.544, 0.307,
  -0.965) to them. Using the control-stream block verbatim instead changes the
  variances by at most 17%.
* **Arm-dependent random effects.** In the authors' control stream the random
  effects on DISP and on the placebo magnitude are applied only to placebo-arm
  patients (`ARM = 0`); in the active arms those two parameters are fixed at
  their covariate-adjusted typical values, and the between-patient variability is
  carried by the treatment effect instead (Results 3.3.1). The model reproduces
  this through the derived indicator `pbo`. The placebo half-life keeps its
  random effect in every arm.
* **Covariate functional forms** are linear and centred (`1 + theta (x - ref)`),
  from the control stream. The BMI effects in the control stream are fixed to 0
  and are therefore omitted. The control stream sets a factor to 1 when a
  covariate is missing (-99); supply complete covariates when simulating.
* **Symptom-duration units.** Table 1 reports the duration of BPH symptoms in
  years (median 4) and the control stream centres it at 4.00, so the coefficient
  is per year; a Supporting Information table note labels it in months, which is
  taken as a slip.
* **Supporting Information worked example.** The text illustrating the baseline
  IPSS effect swaps its two cases: with `DISP x (1 + 0.027 (IPSSb - 16))`, IPSSb
  = 13 gives 0.319 /month and IPSSb = 27 gives 0.450 /month, not the reverse.
* **Abstract versus tables.** The abstract quotes DISP 0.319 /month and
  treatment effects 0.0139 (tamsulosin) and 0.0310 (combination), which differ
  from Tables 2 and 3 (0.347, 0.015, 0.032) and match neither the control stream;
  the tables, which the Results text also quotes, are used.
* **Watchful waiting.** CONDUCT's watchful-waiting arm (control stream `ARM` 1
  and 2) receives its own constant effect from time 0; the timing of the
  protocol-defined switch to tamsulosin is not a model input.
* **Treatment status before baseline.** The control stream switches all effects
  off for records with `TREAT = 0` (off-treatment records). Time 0 of this model
  is the start of randomised treatment, and screening observations were excluded
  from the analysis, so the placebo effect starts at t = 0.
* **Score bounds.** The ODE keeps the score inside (0, 35] on its own, so the
  control stream's truncation of the individual prediction is a no-op; its
  truncation of simulated observations is applied in the simulation above.
* **Stochastic versus typical trajectories.** The treatment-effect variances
  (1.4-1.9 on the log scale) are large, and IPSS responds non-linearly to the
  effect size, so the median simulated trajectory in the active arms runs a few
  points below the typical-patient curve at 48 months. The paper's VPCs
  (Figure 3, Figure S6) were not digitised, so this simulation is not compared
  against the observed trial data here.
* **Residual error** is additive with one SD per arm (Tables 2 and 3); the
  placebo SD was fixed in the final model, the others estimated.
