---
title: "Aciclovir (D'Agate 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aciclovir (D'Agate 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("DAgate_2024_aciclovir"))$reference`
* Description: `r rxode2::rxode(readModelDb("DAgate_2024_aciclovir"))$description`
* Article: <https://doi.org/10.1002/prp2.1193>

# Population

D'Agate 2024 re-analysed pooled aciclovir plasma data from the two clinical
studies reported by Sampson and colleagues. Study 1 was a single-centre,
open-label pharmacokinetic study of infants at 23-42 weeks gestational age and
under 61 days postnatal age with suspected systemic infection; Study 2 was a
multicentre, open-label study of infants at 23-34 weeks gestational age and
under 45 days postnatal age with suspected systemic herpes simplex virus
infection.

Thirty-two infants were enrolled and contributed 92 plasma samples. Nine
samples were excluded before model development as contaminated or drawn during
infusion, leaving **83 samples from 28 infants** in the final data set
(Results, Section 4). Table 1 of the source reports, over the 32 enrolled:
gestational age 30 weeks (range 23-40), post-menstrual age 31 weeks (25-41),
postnatal age 3 days (1-30), birth weight 1295 g (420-4840), weight 1420 g
(373-5720), 17 female / 15 male, 20 White / 11 Black / 1 Asian, and serum
creatinine 0.9 mg/dL (0.3-1.8). One infant received vasopressin, four
dopamine, and seven epinephrine.

The same information is available programmatically via the model's
`population` metadata (`readModelDb("DAgate_2024_aciclovir")()$population`).

# Model structure

Aciclovir disposition is a **one-compartment model with zero-order intravenous
infusion and first-order elimination** (Results, Section 4.1). The final
covariate model (Methods Eq. 3 and the equations printed inside Table 2) is

```
CL (L/h) = theta1 * (WT / 1.37)^0.75 * PMA^HILL / (PMA50^HILL + PMA^HILL) + CLCR
V  (L)   = theta2 * (WT / 1.37)^1 * theta4^DIS
```

Three features are worth spelling out.

* **Clearance is a sum, not a product.** `theta1` is the *residual* clearance
  `theta_CL,res` that creatinine clearance does not explain; the Discussion
  attributes it to tubular secretion plus hepatic metabolism. The individual
  creatinine clearance enters additively at **unit slope**, representing
  glomerular filtration, and the paper reports that it accounts for roughly a
  quarter of total aciclovir clearance.
* **`CLCR` is the individual, non-BSA-normalized value in L/h.** Methods
  Section 2.3 computes the Schwartz estimate
  `CLCR = k * HT / SCr` (`k = 0.33` below 36 weeks gestational age, `0.45`
  otherwise; `HT` in cm, `SCr` in mg/L), which is BSA-normalized, and then
  states that "the values obtained were converted from mL/min/1.73 m^2 to L/h
  using the individual body surface area of the subject". The packaged model
  therefore takes the canonical `CRCL` column as *raw* mL/min (individual, not
  BSA-normalized) and applies the 0.06 mL/min-to-L/h conversion internally.
* **`HILL` is 1 in the final model.** Table 2 prints the final clearance
  equation with the maturation term collapsed to `PMA / (theta3 + PMA)`, and no
  estimate, standard error or bootstrap interval for `HILL` appears anywhere in
  the paper.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/DAgate_2024_aciclovir.R`. The table below
collects them in one place.

| Equation / parameter | Value | Source location |
|---|---|---|
| One compartment, zero-order infusion, first-order elimination | n/a | Results Section 4.1 |
| `lcl` (theta1, residual CL at 1.37 kg) | 0.748 L/h | Table 2, RSE 18%, bootstrap 0.730 (0.429-1.04) |
| `lvc` (theta2, V at 1.37 kg, no infection) | 1.93 L | Table 2, RSE 39%, bootstrap 1.89 (0.823-3.50) |
| `e_wt_cl` (allometric exponent on CL) | 0.75 (fixed) | Methods Section 2.3 item 1; also printed in the Table 2 CL equation |
| `e_wt_vc` (allometric exponent on V) | 1 (fixed) | Methods Section 2.3, "a pre-defined fixed allometric exponent with value 1" |
| `pma_tm50` (theta3, PMA50) | 50 weeks (fixed) | Table 2, footnote b "Fixed to literature value" |
| `pma_hill` (HILL) | 1 (fixed) | Table 2 final-model CL equation printed as `PMA/(theta3 + PMA)` |
| `e_crcl_cl` (CLCR into CL) | unit slope; 0.06 L/h per mL/min | Table 2 CL equation `+ CLCR`; Methods Section 2.3 item 3 for the L/h conversion |
| `e_infect_vc` (theta4, systemic infection on V) | 2.67 | Table 2, RSE 38%, bootstrap 2.74 (1.46-6.43); orientation from the Abstract |
| `etalcl` variance (Omega1) | 0.451 (75.5% CV) | Table 2, RSE 13% |
| `etalvc` variance (Omega2) | 0.646 (95.3% CV) | Table 2, RSE 23% |
| `etalcl`-`etalvc` covariance | 0.442 (rho 82%) | Table 2, RSE 42.3% |
| `propSd` | sqrt(0.143) = 0.378 | Table 2 sigma1 = 0.143 (37.8% CV), RSE 17% |
| Reference weight 1.37 kg, reference PMA context | n/a | Methods, text beneath Eq. 2 and Eq. 3 |

Two arithmetic identities confirm that the Table 2 Omega column holds
**variances on the log scale** rather than standard deviations or CVs: the
table's own footnote c defines the printed percentages as
`sqrt(exp(Omega) - 1) * 100`, and `sqrt(exp(0.451) - 1) = 75.5%` and
`sqrt(exp(0.646) - 1) = 95.3%` reproduce them exactly. The printed correlation
follows from the same column: `0.442 / sqrt(0.451 * 0.646) = 0.819`, i.e. the
82% shown in Table 2.

# Structural check against the paper's own simulated exposures

Table 3 of the source reports steady-state exposure for virtual cohorts of 100
pre-term and 100 term neonates dosed per the summary of product
characteristics, with the dosing interval selected from BSA-normalized
creatinine clearance. Before simulating a cohort, evaluate the packaged model
at Table 3's own **median** covariates for the two cells whose median
creatinine clearance falls inside the reported dosing band: term neonates
(median CLcr 46.2 mL/min/1.73 m^2, i.e. the 25-50 band, 20 mg/kg every 12 h)
and pre-term neonates (median CLcr 23.53, i.e. the 10-25 band, 20 mg/kg every
24 h). Doses are given as 1-hour intravenous infusions.

```{r typical}
mod <- readModelDb("DAgate_2024_aciclovir")

cells <- tibble::tribble(
  ~cohort,    ~tau, ~WT,  ~PAGE, ~BSA, ~CRCL_BSA, ~auc_pub, ~cmax_pub,
  "Term",       12, 3.30, 41.71, 0.23,     46.20,      134,      15.2,
  "Pre-term",   24, 1.18, 29.00, 0.12,     23.53,       70,      12.8
) |>
  # The model consumes the INDIVIDUAL (non-BSA-normalized) creatinine
  # clearance; Table 3 tabulates the BSA-normalized value alongside the
  # individual body surface area, so de-normalize with BSA / 1.73.
  dplyr::mutate(CRCL = CRCL_BSA * BSA / 1.73,
                id = dplyr::row_number(),
                amt = 20 * WT)

make_typical_events <- function(r) {
  dplyr::bind_rows(
    tibble::tibble(id = r$id, time = seq(0, 120 - r$tau, by = r$tau),
                   amt = r$amt, evid = 1L, dur = 1, cmt = "central"),
    tibble::tibble(id = r$id, time = seq(96, 120, by = 0.05),
                   amt = NA_real_, evid = 0L, dur = NA_real_, cmt = "central")
  ) |>
    dplyr::mutate(WT = r$WT, PAGE = r$PAGE, CRCL = r$CRCL,
                  DIS_INFECT_ACTIVE = 0)
}

ev_typical <- dplyr::bind_rows(
  lapply(seq_len(nrow(cells)), function(i) make_typical_events(cells[i, ]))
) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

sim_typical <-
  rxode2::rxSolve(rxode2::zeroRe(mod), events = ev_typical) |>
  as.data.frame()

typ <- sim_typical |>
  dplyr::filter(!is.na(Cc), time >= 96) |>
  dplyr::group_by(id) |>
  dplyr::summarise(
    cl = dplyr::first(cl), vc = dplyr::first(vc), cmax = max(Cc),
    auc24 = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    .groups = "drop"
  ) |>
  dplyr::left_join(cells |> dplyr::select(id, cohort, tau, amt, auc_pub,
                                          cmax_pub), by = "id") |>
  dplyr::mutate(
    daily_dose = amt * 24 / tau,
    mass_balance = auc24 * cl / daily_dose,
    auc_pct  = 100 * (auc24 - auc_pub) / auc_pub,
    cmax_pct = 100 * (cmax - cmax_pub) / cmax_pub
  )

typ |>
  dplyr::select(cohort, tau, cl, vc, auc24, auc_pub, auc_pct, cmax,
                cmax_pub, cmax_pct) |>
  dplyr::rename(
    "Cohort" = cohort, "tau (h)" = tau, "CL (L/h)" = cl, "V (L)" = vc,
    "AUC0-24 model" = auc24, "AUC0-24 paper" = auc_pub, "AUC % diff" = auc_pct,
    "Cmax model" = cmax, "Cmax paper" = cmax_pub, "Cmax % diff" = cmax_pct
  ) |>
  knitr::kable(digits = 2,
               caption = paste("Typical-value steady-state exposure at the",
                               "Table 3 median covariates, against the",
                               "published medians. AUC in mg*h/L, Cmax in",
                               "mg/L."))
```

Two assertions gate this chunk.

The first is an **internal identity** and is therefore held to numerical
tolerance: at steady state the area under the curve over one 24-hour window,
multiplied by the individual clearance, must equal the dose delivered in that
window. Both sides use the same solved profile, so the only difference is
trapezoidal error on a 0.05-hour grid.

```{r typical-assert}
stopifnot(
  # Mass balance: AUC0-24,ss * CL == dose per 24 h. Deterministic; the only
  # error is trapezoidal integration on the 0.05 h observation grid.
  all(abs(typ$mass_balance - 1) < 0.005),
  # Transcription gate against the paper's own Table 3 medians. Deterministic
  # (typical values, no random effects), so the bound is set just outside the
  # realised -4.0% / -1.8% for AUC and -9.8% / +0.9% for Cmax. A
  # mis-transcribed clearance, volume, reference weight or dose moves these by
  # tens of percent.
  max(abs(typ$auc_pct)) < 12,
  max(abs(typ$cmax_pct)) < 20
)
```

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the two neonatal virtual populations of Table 3: 200 pre-term and 200 term
neonates (the per-arm cap for this package's vignettes), each subject's
covariates drawn log-normally with the median and range printed in Table 3.
Body surface area is derived from body weight by geometric scaling
(`BSA = BSA_median * (WT / WT_median)^(2/3)`), which reproduces the Table 3 BSA
ranges from the Table 3 weight ranges to within about 0.01 m^2 in both cohorts.

Between-subject random effects are drawn **in R** from the published
`etalcl`/`etalvc` covariance matrix and carried as data columns, with
`omega = NA` on the solve. `rxSetSeed()` fixes rxode2's own stream per solver
thread and not across thread counts, so drawing the etas here instead makes the
cohort - and therefore every assertion below - identical on any machine.

```{r cohort}
set.seed(20240521)

# Log-scale SD implied by a printed median and range over ~100 subjects,
# treating the range endpoints as approximately the 0.6th and 99.4th
# percentiles (z = 2.5).
log_sd_from_range <- function(lo, hi) log(hi / lo) / 5

# Published IIV covariance (D'Agate 2024 Table 2): Omega1, covariance, Omega2.
omega_pub <- matrix(c(0.451, 0.442,
                      0.442, 0.646), nrow = 2, ncol = 2)

make_cohort <- function(n, label, pma_m, pma_lo, pma_hi,
                        wt_m, wt_lo, wt_hi, bsa_m,
                        crcl_m, crcl_lo, crcl_hi, id_offset = 0L) {
  wt <- wt_m * exp(stats::rnorm(n, 0, log_sd_from_range(wt_lo, wt_hi)))
  etas <- matrix(stats::rnorm(2 * n), ncol = 2) %*% chol(omega_pub)
  tibble::tibble(
    id       = id_offset + seq_len(n),
    cohort   = label,
    PAGE     = pma_m * exp(stats::rnorm(n, 0,
                                        log_sd_from_range(pma_lo, pma_hi))),
    WT       = wt,
    BSA      = bsa_m * (wt / wt_m)^(2 / 3),
    CRCL_BSA = crcl_m * exp(stats::rnorm(n, 0,
                                         log_sd_from_range(crcl_lo, crcl_hi))),
    etalcl   = etas[, 1],
    etalvc   = etas[, 2]
  ) |>
    dplyr::mutate(
      CRCL = CRCL_BSA * BSA / 1.73,
      # Table 3 simulated the general neonatal population rather than the
      # infected sub-study; see "Assumptions and deviations".
      DIS_INFECT_ACTIVE = 0,
      # Renal dose adjustment exactly as listed in Methods Section 2.5, keyed
      # on BSA-normalized CLcr for neonates.
      tier = dplyr::case_when(CRCL_BSA > 50 ~ "CLcr > 50",
                              CRCL_BSA > 25 ~ "CLcr 25-50",
                              CRCL_BSA > 10 ~ "CLcr 10-25",
                              TRUE          ~ "CLcr < 10"),
      tau  = dplyr::case_when(CRCL_BSA > 50 ~ 8,
                              CRCL_BSA > 25 ~ 12,
                              TRUE          ~ 24),
      amt  = dplyr::if_else(CRCL_BSA > 10, 20, 10) * WT,
      arm  = paste(cohort, tier, sep = ", ")
    )
}

subj <- dplyr::bind_rows(
  make_cohort(200, "Pre-term", 29.00, 24.22, 38.73, 1.18, 0.32, 5.59, 0.12,
              23.53, 15.18, 49.89, id_offset =   0L),
  make_cohort(200, "Term",     41.71, 37.54, 45.30, 3.30, 1.28, 8.38, 0.23,
              46.20, 27.01, 89.71, id_offset = 200L)
)
stopifnot(!anyDuplicated(subj$id))

subj |>
  dplyr::count(cohort, tier) |>
  tidyr::pivot_wider(names_from = tier, values_from = n, values_fill = 0L) |>
  knitr::kable(caption = paste("Subjects per renal dosing band. The pattern of",
                               "populated cells matches Table 3, where the",
                               "pre-term > 50 and term 10-25 cells are",
                               "reported as absent."))
```

The dosing-band occupancy is itself a check on the covariate model: applying
the paper's own creatinine-clearance dose rules to the paper's own creatinine
clearance distributions leaves the pre-term `CLcr > 50` and term
`CLcr 10-25` cells empty, exactly the two cells Table 3 leaves blank for these
cohorts.

```{r events}
t_ss <- 96  # steady state is reached within about a day (t1/2 ~ 3 h)

dose_rows <- subj |>
  dplyr::group_by(id) |>
  dplyr::reframe(time = seq(0, t_ss + 24 - unique(tau), by = unique(tau)),
                 amt = unique(amt), evid = 1L, dur = 1, cmt = "central")

obs_rows <- subj |>
  dplyr::select(id) |>
  tidyr::crossing(time = seq(t_ss, t_ss + 24, by = 0.25)) |>
  dplyr::mutate(amt = NA_real_, evid = 0L, dur = NA_real_, cmt = "central")

events <- dplyr::bind_rows(dose_rows, obs_rows) |>
  dplyr::left_join(
    subj |> dplyr::select(id, arm, cohort, WT, PAGE, CRCL, DIS_INFECT_ACTIVE,
                          etalcl, etalvc),
    by = "id"
  ) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(mod, events = events, omega = NA, sigma = NA,
                       keep = c("arm", "cohort")) |>
  as.data.frame()

stopifnot(all(sim$Cc[!is.na(sim$Cc)] >= 0))
```

`omega = NA` and `sigma = NA` suppress rxode2's own sampling: the between-
subject variability is already carried in the `etalcl` / `etalvc` columns, and
`Cc` is the individual prediction. Table 3's secondary parameters are
model-predicted exposures, so no residual error is added.

# Replicate published figures

```{r figure-3}
# Replicates the neonatal panels of Figure 3 of D'Agate 2024: steady-state
# aciclovir concentration versus time within one 24-hour window.
sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(time_ss = time - t_ss) |>
  dplyr::group_by(arm, time_ss) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot2::ggplot(ggplot2::aes(time_ss, Q50)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  ggplot2::geom_line() +
  ggplot2::facet_wrap(~arm) +
  ggplot2::scale_y_log10() +
  ggplot2::labs(
    x = "Time within the steady-state dosing day (h)",
    y = "Aciclovir plasma concentration (mg/L)",
    title = "Steady-state aciclovir profiles in pre-term and term neonates",
    caption = paste("Replicates the neonatal panels of Figure 3 of D'Agate",
                    "2024. Line = median, band = 5th-95th percentile.")
  )
```

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(sim_nca), Cc ~ time | arm + id)

dose_df <- events |>
  dplyr::filter(evid == 1, time >= t_ss) |>
  dplyr::select(id, time, amt, arm)

dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | arm + id)

# The steady-state 24-hour window. Every subject has an observation exactly at
# t_ss, so the AUC interval is anchored on a real measurement.
intervals <- data.frame(
  start   = t_ss,
  end     = t_ss + 24,
  cmax    = TRUE,
  tmax    = TRUE,
  auclast = TRUE,
  cav     = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)

nca_tbl <- as.data.frame(nca_res$result)
stopifnot(nrow(nca_tbl) > 0, !anyNA(nca_tbl$PPORRES))
```

```{r pknca-summary}
nca_tbl |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "cav")) |>
  dplyr::group_by(arm, PPTESTCD) |>
  dplyr::summarise(n = dplyr::n(),
                   Median = median(PPORRES),
                   `P5` = quantile(PPORRES, 0.05),
                   `P95` = quantile(PPORRES, 0.95),
                   .groups = "drop") |>
  dplyr::rename("Arm" = arm, "NCA parameter" = PPTESTCD, "N" = n) |>
  knitr::kable(digits = 2,
               caption = paste("Simulated steady-state NCA by renal dosing",
                               "band. AUClast is AUC0-24 at steady state",
                               "(mg*h/L); Cmax and Cav in mg/L; Tmax in h."))
```

## Comparison against published NCA

```{r nca_compare}
published <- tibble::tribble(
  ~arm,                   ~auclast, ~cmax,
  "Term, CLcr > 50",           185,  12.7,
  "Term, CLcr 25-50",          134,  15.2,
  "Pre-term, CLcr 25-50",      134,  14.8,
  "Pre-term, CLcr 10-25",       70,  12.8
)

# Guard against a silent no-op: every published arm must exist in the
# simulation, otherwise the comparison table would render empty rows and the
# assertions below would test nothing.
stopifnot(all(published$arm %in% nca_tbl$arm))

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "arm",
  units         = c(cmax = "mg/L", auclast = "mg*h/L", tmax = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste("Simulated versus published steady-state exposure",
                  "(D'Agate 2024 Table 3). * differs from the reference by",
                  "more than 20%."),
  align = c("l", "l", "r", "r", "r")
)
```

```{r nca_assert}
pct_diff <- nca_tbl |>
  dplyr::filter(PPTESTCD %in% c("auclast", "cmax")) |>
  dplyr::group_by(arm, PPTESTCD) |>
  dplyr::summarise(median_sim = median(PPORRES), .groups = "drop") |>
  dplyr::left_join(
    published |> tidyr::pivot_longer(-arm, names_to = "PPTESTCD",
                                     values_to = "published"),
    by = c("arm", "PPTESTCD")
  ) |>
  dplyr::mutate(pct = 100 * (median_sim - published) / published)

stopifnot(nrow(pct_diff) == 8L, !anyNA(pct_diff$pct))

auc_pct  <- pct_diff$pct[pct_diff$PPTESTCD == "auclast"]
# The term / CLcr > 50 Cmax cell is a documented deviation -- see
# "Assumptions and deviations" -- and is excluded from the gate but kept
# visible in the comparison table above.
cmax_pct <- pct_diff$pct[pct_diff$PPTESTCD == "cmax" &
                           pct_diff$arm != "Term, CLcr > 50"]

stopifnot(
  # Realised -19.7 / -5.7 / +0.6 / +10.4 for AUC0-24 and -16.4 / -4.6 / +10.1
  # for the three gated Cmax cells. The cohort is drawn from R's RNG under
  # set.seed() with rxode2 sampling disabled, so these are reproducible; the
  # bound leaves headroom for solver and PKNCA-version differences while
  # still breaking on a mis-transcribed clearance, volume or dose.
  max(abs(auc_pct))  < 28,
  max(abs(cmax_pct)) < 25
)
```

The four AUC0-24 cells agree with the published medians to within 20%, and
three of the four Cmax cells to within 17%.

# Assumptions and deviations

* **Volume-of-distribution covariate form.** Table 2 prints the volume equation
  as `V (L) = theta2*(WT/1.37)*DIS*theta4`, which taken literally sets the
  volume to zero for a subject without systemic infection. The Abstract states
  the intended form: "Population estimate for volume of distribution was 1.93 L
  with systemic infection increasing this value by almost 3-fold (2.67 times
  higher)". The model therefore encodes `theta4^DIS`, i.e. a 2.67-fold
  multiplier when `DIS_INFECT_ACTIVE = 1` and no change when it is 0.

* **`HILL` is not reported.** The Methods Eq. 3 maturation term carries a Hill
  coefficient, but no estimate for it appears in Table 2 or anywhere else in
  the paper. Table 2's own printed final-model equation collapses the term to
  `PMA / (theta3 + PMA)`, so `pma_hill` is encoded as `fixed(1)`.

* **Systemic infection is switched off in the simulations.** The paper does not
  state which value of `DIS` its Table 3 virtual cohorts carry. Setting
  `DIS_INFECT_ACTIVE = 0` reproduces the published Cmax values; setting it to 1
  divides every simulated Cmax by 2.67 and puts them far below the published
  numbers (for example the term median falls from about 13 mg/L to about
  5 mg/L). The Table 3 cohorts were also drawn from the NHANES and CALIPER
  reference databases rather than from the infected study population, so
  `DIS = 0` is both the numerically and the clinically consistent reading.

* **Infusion duration 1 hour.** Neither the Methods nor Table 3 states the
  infusion duration used for the simulations. One hour is the duration in the
  summary of product characteristics the paper cites for the dosing regimens,
  and it is corroborated numerically: the typical-value term neonate at the
  Table 3 median covariates reaches a steady-state Cmax of 13.7 mg/L on a
  1-hour infusion against the published 15.2 mg/L, whereas a bolus would
  overshoot it.

* **Covariate distributions are reconstructed, not published.** Table 3 gives a
  median and range per covariate but no distributional form or correlation
  structure. Post-menstrual age, weight and BSA-normalized creatinine clearance
  are drawn log-normally with the log-scale SD implied by treating the printed
  range endpoints as roughly the 0.6th and 99.4th percentiles of 100 subjects,
  and body surface area is derived from weight by geometric (2/3-power)
  scaling anchored at the published cohort medians. That derivation is not
  taken from the paper, but it reproduces the Table 3 BSA ranges from the
  Table 3 weight ranges in both cohorts, which is why it is preferred here to
  sampling BSA independently of weight.

* **Between-subject random effects are drawn in R.** They are passed to
  `rxSolve()` as `etalcl` / `etalvc` data columns with `omega = NA`, rather
  than sampled by rxode2. rxode2 partitions its RNG streams per solver thread,
  so a thread-count difference between a workstation and a CI runner would
  otherwise draw a different cohort and move every assertion above.

* **Known deviation: Cmax in the term, CLcr > 50 cell.** The simulated median
  Cmax is about 26% above the published 12.7 mg/L. The source of the
  disagreement is visible inside Table 3 itself: at a constant 20 mg/kg unit
  dose, the table reports a *lower* Cmax for the more frequent regimen
  (term neonates 12.7 mg/L every 8 h against 15.2 mg/L every 12 h; infants
  14.7 mg/L against 17.5 mg/L). A one-compartment model with first-order
  elimination accumulates more, not less, as the interval shortens at constant
  unit dose, so this ordering cannot arise from the published structural model
  regardless of how the covariates are distributed. The corresponding AUC0-24
  cell agrees to within 6%, so the transcription of clearance, volume and dose
  is not implicated. The cell is reported in the comparison table and excluded
  from the assertion gate rather than tuned away.

* **Covariates screened but not retained.** Post-natal age, sex, race,
  concomitant vasopressin / dopamine / epinephrine, birth weight and serum
  creatinine as a covariate in its own right were all tested and dropped
  (Methods Section 2.2 and Discussion). They are recorded in the model file's
  `covariatesDataExcluded` metadata so the covariate screen is preserved
  without implying an effect the paper did not estimate.
