---
title: "Dalteparin (Damle 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dalteparin (Damle 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
suppressPackageStartupMessages({
  library(nlmixr2lib)
  library(PKNCA)
  library(rxode2)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Damle_2021_dalteparin"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Damle_2021_dalteparin"))$description`
* Article: <https://doi.org/10.1002/jcph.1716> (open access)

Dalteparin is a low-molecular-weight heparin. Because the drug itself is
difficult to assay, its pharmacokinetics are described through plasma
anti-factor Xa (anti-Xa) activity, so the model's "concentration" `Cc` is
anti-Xa activity in IU/mL and doses are in anti-Xa IU.

# Population

Damle 2021 pooled 266 anti-Xa observations from 89 pediatric patients with
acute venous thromboembolism (VTE) from three sources: a Pfizer-sponsored
prospective open-label study (NCT00952380, 15 sites in North America and
Europe, n = 37), the dose-finding pilot of the Kids-DOTT trial (n = 18) and a
Mayo Clinic retrospective chart review (n = 34). Age ranged from 15 days to
19.5 years (median 12.0 years), body weight from 2.3 to 161 kg (median
43.4 kg); 66.3% were male, 45% were white and 48.3% had cancer (Table 1).
Patients received subcutaneous dalteparin twice daily, starting at
100-150 IU/kg by age group and titrated to a 4-6 h post-dose anti-Xa target of
0.5-1.0 IU/mL; most samples were drawn 3-6 h after a dose.

```{r population}
str(rxode2::rxode2(readModelDb("Damle_2021_dalteparin"))$population)
```

# Source trace

| Model element | Value | Source location |
|---|---|---|
| Structure: 1-compartment, first-order absorption and elimination, parameterised in CL/F, V/F, ka | - | Results, Population Pharmacokinetic Analysis |
| `lcl` (CL/F, 43 kg, 12 y, female, with cancer) | 929 mL/h = 0.929 L/h | Table 2, theta1 |
| `lvc` (V/F, 43 kg) | 7180 mL = 7.18 L | Table 2, theta2 |
| `lka` | 1.04 1/h | Table 2, theta3 |
| `vc_eta_scale` | 1.73 | Table 2, theta5; equation eta_V = theta5 * eta_CL |
| `e_wt_cl` | 0.75 (fixed) | Table 2, theta6; Methods |
| `e_wt_vc` | 1 (fixed) | Table 2, theta7; Methods |
| `e_age_cl` | -0.0687 | Table 2, theta8 |
| `e_sexm_cl` | 1.03 (male) | Table 2, theta14 |
| `e_nocancer_cl` | 0.885 (without cancer) | Table 2, theta15 |
| `etalcl` | 0.0369 (19% CV) | Table 2, omega2 CL/F |
| `propSd` | sqrt(0.0519) = 0.228 | Table 2, sigma2 P (text: 23% CV) |
| `addSd` | sqrt(0.016) = 0.1265 IU/mL | Table 2, sigma2 A |
| TVCL = theta1 (WT/43)^theta6 (AGE/12)^theta8 theta14^I(male) theta15^I(no cancer) | - | Results, model equations (page 175) |
| TVV = theta2 (WT/43)^theta7 | - | Results, model equations (page 175) |

# Virtual cohort

The paper simulated 1000 subjects per age group "based on the age-weight
relationship in the current population pharmacokinetic data set", which is
not published. Here age is drawn uniformly within each of the paper's five
age groups, and body weight is drawn log-normally (20% CV) around a weight-age
curve interpolated through the per-group median ages and weights of Table 1.
Each group has 200 subjects. The between-subject random effect `etalcl` is
drawn with base R and passed to the model as a data column, so the cohort does
not depend on the rxode2 random-number stream and the checks below are
deterministic.

```{r cohort}
mod <- rxode2::rxode2(readModelDb("Damle_2021_dalteparin"))
omega_cl <- mod$omega["etalcl", "etalcl"]
prop_sd <- mod$theta[["propSd"]]
add_sd <- mod$theta[["addSd"]]

# Table 1 per-group median age (years) and weight (kg)
anchor <- data.frame(
  age = c(0.06, 0.5, 4.5, 9.6, 15.9),
  wt = c(3.5, 6.8, 14.6, 36.1, 61.8)
)
median_wt <- function(age) stats::approx(anchor$age, anchor$wt, xout = age, rule = 2)$y

groups <- data.frame(
  group = c(
    "0 to <8 weeks", ">=8 weeks to <2 years", ">=2 to <8 years",
    ">=8 to <12 years", ">=12 to <19 years"
  ),
  age_lo = c(15 / 365.25, 8 / 52, 2, 8, 12),
  age_hi = c(8 / 52, 2, 8, 12, 19)
)
groups$group <- factor(groups$group, levels = groups$group)

n_per_group <- 200L # cohort cap: never more than 200 participants per arm
set.seed(20210201)
cohort <- bind_rows(lapply(seq_len(nrow(groups)), function(i) {
  age <- stats::runif(n_per_group, groups$age_lo[i], groups$age_hi[i])
  data.frame(
    group = groups$group[i],
    AGE = age,
    WT = median_wt(age) * exp(stats::rnorm(n_per_group, 0, 0.2)),
    etalcl = stats::rnorm(n_per_group, 0, sqrt(omega_cl))
  )
})) |>
  mutate(
    id = row_number(),
    # The paper's dose simulations used a reduced model without the sex and
    # cancer factors; with theta14 = theta15 = 1 dropped that is the full model
    # at its reference levels (female, with cancer).
    SEXF = 1,
    DIS_CANCER_PED = 1
  )

cohort |>
  group_by(group) |>
  summarise(
    n = n(),
    `Age median (y)` = signif(median(AGE), 3),
    `WT median (kg)` = signif(median(WT), 3),
    `WT range (kg)` = paste(signif(range(WT), 3), collapse = "-")
  ) |>
  knitr::kable()
```

# Simulation

Steady-state twice-daily (every 12 h) dosing is simulated with `ss = 1`, and
observations are taken on the `central` state. The model is solved with its
random effects zeroed (`rxode2::zeroRe()`) because the per-subject `etalcl`
comes from the data; the residual error is then applied analytically below.

```{r sim-helpers}
mod_typ <- rxode2::zeroRe(mod)

# The random-effect variances are zeroed on purpose, so the "simulation
# without omega" warning is expected and muffled; any other warning surfaces.
solve_ss <- function(events) {
  withCallingHandlers(
    rxode2::rxSolve(mod_typ, events, returnType = "data.frame", keep = "group"),
    warning = function(w) {
      if (grepl("without 'omega'", conditionMessage(w), fixed = TRUE)) {
        invokeRestart("muffleWarning")
      }
    }
  )
}

make_events <- function(cohort, dose_per_kg, obs_times) {
  dose_rows <- cohort |>
    mutate(
      time = 0, amt = dose_per_kg * WT, evid = 1L, cmt = "depot",
      ss = 1L, ii = 12
    )
  obs_rows <- tidyr::crossing(cohort, time = obs_times) |>
    mutate(amt = 0, evid = 0L, cmt = "central", ss = 0L, ii = 0)
  bind_rows(dose_rows, obs_rows) |>
    arrange(id, time, desc(evid))
}
```

## Probability of target attainment (Figure 4)

For each age group and each dose from 75 to 300 IU/kg every 12 h, the
steady-state anti-Xa level 4 h after a dose (C4hss) is computed. The
probabilities of target attainment (0.5-1.0 IU/mL), underattainment and
overattainment then integrate the combined residual error of the model over
each subject's individual prediction: an observation is normal around the
prediction with SD `sqrt(addSd^2 + (propSd * pred)^2)`. Including the residual
error is what reproduces the paper's figure. Using the individual prediction
alone gives maximum PTAs near 0.85, where the paper reports 0.61-0.66.

```{r pta}
doses <- seq(75, 300, by = 25)
c4 <- bind_rows(lapply(doses, function(d) {
  solve_ss(make_events(cohort, d, obs_times = 4)) |>
    filter(time == 4) |>
    transmute(id, group, dose = d, pred = Cc)
}))

pta <- c4 |>
  mutate(
    sd = sqrt(add_sd^2 + (prop_sd * pred)^2),
    p_under = stats::pnorm(0.5, pred, sd),
    p_over = 1 - stats::pnorm(1, pred, sd)
  ) |>
  group_by(group, dose) |>
  summarise(
    under = mean(p_under),
    over = mean(p_over),
    .groups = "drop"
  ) |>
  mutate(pta = 1 - under - over)
```

Values digitised by the maintainers from Figure 4 of Damle 2021 (read to about
+/-0.02) are overlaid as points.

```{r fig4-digitised}
fig4 <- tibble::tibble(
  group = rep(levels(groups$group), each = length(doses)),
  dose = rep(doses, times = 5),
  pta = c(
    0.01, 0.04, 0.11, 0.20, 0.30, 0.45, 0.56, 0.61, 0.66, 0.66,
    0.07, 0.18, 0.33, 0.49, 0.58, 0.64, 0.62, 0.61, 0.55, 0.47,
    0.15, 0.33, 0.52, 0.63, 0.64, 0.61, 0.51, 0.40, 0.33, 0.25,
    0.27, 0.53, 0.61, 0.60, 0.54, 0.44, 0.33, 0.25, 0.15, 0.12,
    0.32, 0.54, 0.64, 0.58, 0.48, 0.36, 0.27, 0.17, 0.13, 0.09
  ),
  over = c(
    0.00, 0.00, 0.00, 0.00, 0.005, 0.01, 0.02, 0.04, 0.10, 0.15,
    0.00, 0.00, 0.005, 0.02, 0.06, 0.10, 0.20, 0.27, 0.38, 0.48,
    0.00, 0.00, 0.03, 0.09, 0.18, 0.30, 0.46, 0.55, 0.65, 0.73,
    0.005, 0.02, 0.10, 0.24, 0.38, 0.51, 0.64, 0.74, 0.84, 0.87,
    0.01, 0.05, 0.13, 0.30, 0.46, 0.61, 0.72, 0.82, 0.86, 0.91
  )
) |>
  mutate(group = factor(group, levels = levels(groups$group)))
```

```{r fig4, fig.height = 6}
pta_long <- pta |>
  select(group, dose, `Target (0.5-1.0 IU/mL)` = pta, `Over (>1.0 IU/mL)` = over, `Under (<0.5 IU/mL)` = under) |>
  pivot_longer(-c(group, dose), names_to = "outcome", values_to = "probability")
fig4_long <- fig4 |>
  select(group, dose, `Target (0.5-1.0 IU/mL)` = pta, `Over (>1.0 IU/mL)` = over) |>
  pivot_longer(-c(group, dose), names_to = "outcome", values_to = "probability")

ggplot(pta_long, aes(dose, probability, colour = outcome)) +
  geom_line(linewidth = 0.9) +
  geom_point(data = fig4_long, shape = 1, size = 2) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  facet_wrap(~group) +
  scale_colour_manual(values = c(
    "Target (0.5-1.0 IU/mL)" = "forestgreen",
    "Over (>1.0 IU/mL)" = "red3", "Under (<0.5 IU/mL)" = "orange"
  )) +
  labs(
    x = "Dalteparin dose every 12 h (IU/kg)", y = "Probability",
    colour = NULL,
    caption = "Replicates Figure 4 of Damle 2021. Lines: this model; open points: digitised from the paper."
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

The dose at which the rising PTA curve crosses 0.5 is the paper's criterion
for the starting dose (dashed lines in its Figure 4; the paper rounds these to
200, 150, 125, 100 and 100 IU/kg).

```{r pta-check}
dose_at_half <- function(dose, p) {
  i <- which(p >= 0.5)[1]
  stats::approx(p[(i - 1):i], dose[(i - 1):i], xout = 0.5)$y
}
pta_summary <- pta |>
  group_by(group) |>
  summarise(
    dose50 = dose_at_half(dose, pta),
    max_pta = max(pta),
    .groups = "drop"
  ) |>
  mutate(
    paper_dose50 = c(210, 155, 122, 97, 95),
    paper_max_pta = fig4 |> group_by(group) |> summarise(m = max(pta)) |> pull(m),
    dose50_pct_diff = 100 * (dose50 - paper_dose50) / paper_dose50
  )
pta_summary |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  rename(
    `Age group` = group,
    `Dose at 50% PTA, model (IU/kg)` = dose50,
    `Max PTA, model` = max_pta,
    `Dose at 50% PTA, Figure 4 (IU/kg)` = paper_dose50,
    `Max PTA, Figure 4` = paper_max_pta,
    `Dose difference (%)` = dose50_pct_diff
  ) |>
  knitr::kable()

pta_diff <- pta |>
  inner_join(fig4, by = c("group", "dose"), suffix = c("", "_paper")) |>
  mutate(abs_diff = abs(pta - pta_paper))

stopifnot(
  # Per-group 50%-PTA dose within 15% of the digitised dashed line: a 20%
  # error in CL/F moves this dose by ~20%.
  all(abs(pta_summary$dose50_pct_diff) < 15),
  # Height of the bell curve: the paper reports 0.608-0.662.
  all(abs(pta_summary$max_pta - pta_summary$paper_max_pta) < 0.08),
  # Whole-curve agreement over the 50 digitised PTA points.
  median(pta_diff$abs_diff) < 0.05,
  quantile(pta_diff$abs_diff, 0.9) < 0.12
)
```

The bell-shaped PTA curves and their maximum heights (about 0.65) match the
paper. For the four groups younger than 12 years the model's 50%-PTA dose sits
about 8-9% above the paper's dashed lines; the adolescent group matches. Two
unpublished inputs could explain the small offset: the age-weight relationship
of the analysis data set, which the paper used to generate its cohort, and
whether the reduced model was re-estimated after the sex and cancer factors
were dropped. Rounded to the paper's 25 IU/kg grid, the model reproduces the
recommended starting doses of 125 IU/kg (2 to <8 years) and 100 IU/kg
(8 to <19 years). It gives 175 rather than 150 IU/kg for 8 weeks to <2 years
and 225 rather than 200 IU/kg for the youngest group; for that group the paper
itself set the starting dose conservatively below its 50%-PTA dose.

## Weight-normalised clearance versus age (Figure 3)

Figure 3 of the paper plots the individual CL/F per kg against age with a
smoother. Typical CL/F per kg is computed here along the Table 1 weight-age
curve and compared with points read by the maintainers from the paper's
smoothed line.

```{r fig3}
theta <- mod$theta
typ_cl_per_kg <- function(age) {
  wt <- median_wt(age)
  1000 * exp(theta[["lcl"]]) * (wt / 43)^theta[["e_wt_cl"]] *
    (age / 12)^theta[["e_age_cl"]] / wt
}
curve <- data.frame(age = seq(0.05, 19.5, by = 0.05)) |>
  mutate(cl_per_kg = typ_cl_per_kg(age))
fig3 <- data.frame(
  age = c(2.5, 5, 10, 15, 19.5),
  cl_per_kg = c(35, 27, 21.5, 19.3, 17)
)

ggplot(curve, aes(age, cl_per_kg)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(data = fig3, shape = 1, size = 3) +
  labs(
    x = "Age (years)", y = "Typical CL/F per kg (mL/h/kg)",
    caption = "Replicates Figure 3 of Damle 2021. Line: typical value; open points: the paper's smoother, digitised."
  ) +
  theme_bw()

fig3_check <- fig3 |>
  mutate(
    model = typ_cl_per_kg(age),
    pct_diff = 100 * (model - cl_per_kg) / cl_per_kg
  )
knitr::kable(fig3_check, digits = 1)
stopifnot(all(abs(fig3_check$pct_diff) < 15))
```

At the reference patient (12 years, 43 kg) the typical value is
929 / 43 = 21.6 mL/h/kg.

# PKNCA validation

The paper reports no noncompartmental parameters, so the NCA is checked
against the closed-form steady-state solution of the one-compartment
first-order-absorption model, using each subject's own CL/F and V/F. Each
age group receives the starting dose the paper recommends (150 IU/kg for
patients under 2 years, 125 IU/kg for 2 to <8 years and 100 IU/kg from
8 years) every 12 h to steady state.

```{r nca-sim}
start_dose <- c(150, 150, 125, 100, 100)
names(start_dose) <- levels(groups$group)
obs_grid <- seq(0, 12, by = 0.1)

ss_prof <- bind_rows(lapply(levels(groups$group), function(g) {
  sub <- filter(cohort, group == g)
  solve_ss(make_events(sub, start_dose[[g]], obs_times = obs_grid)) |>
    mutate(dose_per_kg = start_dose[[g]])
})) |>
  mutate(amt = dose_per_kg * WT)

ss_conc <- ss_prof |>
  filter(!is.na(Cc)) |>
  select(id, group, time, Cc)
ss_dose <- ss_prof |>
  distinct(id, group, amt) |>
  mutate(time = 0)

conc_obj <- PKNCA::PKNCAconc(ss_conc, Cc ~ time | group + id)
dose_obj <- PKNCA::PKNCAdose(ss_dose, amt ~ time | group + id)
intervals <- data.frame(start = 0, end = 12, auclast = TRUE, cmax = TRUE, tmax = TRUE, cmin = TRUE)
nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_res <- as.data.frame(nca$result)
```

```{r nca-summary}
nca_res |>
  select(group, id, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  group_by(group) |>
  summarise(
    across(c(cmax, tmax, cmin, auclast), ~ signif(median(.x), 3))
  ) |>
  rename(
    `Age group` = group,
    `Cmax,ss (IU/mL)` = cmax,
    `Tmax (h)` = tmax,
    `Cmin,ss (IU/mL)` = cmin,
    `AUC0-12,ss (IU*h/mL)` = auclast
  ) |>
  knitr::kable()
```

At steady state `AUC(0-12) = Dose / (CL/F)`, and the 4 h concentration follows
the closed-form one-compartment equation. Both sides use the same individual
parameters, so the only difference is the trapezoidal and ODE-solver error,
and a tight bound applies.

```{r nca-closed-form}
ind <- ss_prof |>
  filter(time == 4) |>
  select(id, group, amt, cl, vc, ka, C4 = Cc)
closed <- nca_res |>
  filter(PPTESTCD == "auclast") |>
  select(id, auc = PPORRES) |>
  inner_join(ind, by = "id") |>
  mutate(
    kel = cl / vc,
    auc_closed = amt / (cl * 1000),
    c4_closed = amt / 1000 * ka / (vc * (ka - kel)) *
      (exp(-kel * 4) / (1 - exp(-kel * 12)) - exp(-ka * 4) / (1 - exp(-ka * 12))),
    auc_pct = 100 * (auc - auc_closed) / auc_closed,
    c4_pct = 100 * (C4 - c4_closed) / c4_closed
  )
closed |>
  group_by(group) |>
  summarise(
    `Max abs AUC diff (%)` = signif(max(abs(auc_pct)), 2),
    `Max abs C4h diff (%)` = signif(max(abs(c4_pct)), 2)
  ) |>
  rename(`Age group` = group) |>
  knitr::kable()
stopifnot(
  nrow(closed) == nrow(cohort),
  all(abs(closed$auc_pct) < 0.5),
  all(abs(closed$c4_pct) < 0.1)
)
```

# Assumptions and deviations

* **Additive residual error.** Table 2 reports both residual terms as
  variances (sigma2 P = 0.0519, sigma2 A = 0.016). The text converts the
  proportional term to its SD (23% CV = sqrt(0.0519)) but quotes the additive
  term as "0.016 IU/mL", which is the variance. The model uses SDs for both:
  `propSd = sqrt(0.0519) = 0.228` and `addSd = sqrt(0.016) = 0.1265 IU/mL`.
  Both terms enter as independent errors (proportional plus additive).
* **Full model, not the reduced model.** The model file encodes the full
  covariate model of Table 2, the only parameter set the paper prints. The
  paper's dose simulations used a reduced model with the sex and cancer factors
  dropped; no separate estimates are reported for it. The simulations above
  therefore use the full model at its reference levels (female, with cancer),
  where both factors equal 1. The close agreement with Figure 4 is consistent
  with the reduced model not having been re-estimated, but the paper does not
  say.
* **Covariate coding.** The paper codes SEX = 1 for male and CANCERST = 1 for
  patients without cancer, and applies each factor when its indicator is 1.
  The model uses the canonical `SEXF` (1 = female) and `DIS_CANCER_PED`
  (1 = with cancer) columns, with the factors applied as
  `e_sexm_cl^(1 - SEXF)` and `e_nocancer_cl^(1 - DIS_CANCER_PED)`. The
  reference patient is unchanged: a 12-year-old, 43 kg female with cancer.
* **Correlated IIV.** The paper estimates eta_V/F = theta5 x eta_CL/F, i.e.
  a correlation of exactly 1. This is encoded as `vc_eta_scale * etalcl`,
  because a correlation-1 OMEGA block is singular. The implied V/F variability
  is 1.73 x 19.2% = 33% CV, matching the text. There is no IIV on ka (the
  paper removed it because few samples were drawn in the absorption phase).
* **Unit conversion.** CL/F and V/F are converted from mL to L, and `Cc`
  divides by 1000 so anti-Xa activity is in IU/mL.
* **Virtual cohort.** The paper's age-weight relationship is not published.
  The weight-age curve here interpolates the Table 1 group medians, with 20% CV
  log-normal scatter, 200 subjects per age group (the paper used 1000). The
  Table 1 weight range for the 2 to <8 years group (11.5-161 kg) includes a
  value that is implausible for that age and is not used.
* **Digitised values.** The Figure 4 PTA and overattainment values and the
  Figure 3 smoother points were read by the maintainers from the published
  figures and are accurate to roughly +/-0.02 (probability) and +/-1 mL/h/kg.
* **Bioavailability.** Only subcutaneous dosing was studied, so F is not
  identifiable. CL/F and V/F are apparent values, and doses go to `depot`.
* **Errata.** None found: Crossref lists no correction or update for
  doi:10.1002/jcph.1716, and the article has no supplementary data beyond its
  figures (checked 2026-09-27).
