---
title: "Treosulfan (Danielak 2017)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Treosulfan (Danielak 2017)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Danielak D, Twardosz J, Kasprzyk A, Wachowiak J, Kalwak K,
  Glowka F. Population pharmacokinetics of treosulfan and development of
  a limited sampling strategy in children prior to hematopoietic stem
  cell transplantation. *Eur J Clin Pharmacol*. 2018 Jan;74(1):79-89.
* DOI: <https://doi.org/10.1007/s00228-017-2344-x>
* Model description: `r nlmixr2lib::modeldb$description[nlmixr2lib::modeldb$name == "Danielak_2017_treosulfan"]`

# Population

The model was developed from data collected 2007-2011 at the Department of
Oncology, Hematology and Pediatric Transplantation (Poznan University of
Medical Sciences) and the Department of Pediatric Hematology, Oncology and
Bone Marrow Transplantation (Wroclaw Medical University). Fifteen pediatric
patients (12 boys, 3 girls; age 7.8 +/- 4.9 years, range 0.4-15; body
weight 26.9 +/- 15.7 kg, range 7.7-52; body surface area 0.95 +/- 0.44
m^2) received treosulfan-based conditioning prior to allogeneic
hematopoietic stem cell transplantation. Diagnoses spanned hematologic
malignancies (ALL 4, AML 1, CML 1), solid tumors (neuroblastoma 2, Ewing
sarcoma 2), and non-malignant disorders (adrenoleukodystrophy 2,
Diamond-Blackfan anemia 1, severe congenital neutropenia 1, Wiskott-Aldrich
syndrome 1). Treosulfan was administered IV at 10, 12, or 14 g/m^2 once
daily as a 1 h or 2 h infusion. A total of 110 plasma concentrations were
collected on day 1 of therapy across two sparse-sampling protocols (6
timepoints in 7 subjects; 10 timepoints in 8 subjects); all observed
concentrations were above the HPLC-MS/MS LOQ of 0.2 uM. Creatinine
clearance (available for 8 of 15 patients) and sex were tested and not
retained as covariates; only body weight survived backward elimination
(Danielak 2017 Methods + Table 1).

The full demographics list, including age strata, dosing regimens, and
disease state breakdown, is encoded in the model file under the
`population` metadata block (Danielak 2017 Table 1).

# Source trace

The per-parameter origin is also recorded as an in-file comment next to
each `ini()` entry in
`inst/modeldb/specificDrugs/Danielak_2017_treosulfan.R`. The table below
consolidates them.

| Equation / parameter      | Value                | Source location                                              |
|---------------------------|----------------------|--------------------------------------------------------------|
| `lcl` (CL at 70 kg)       | log(14.7 L/h)        | Danielak 2017 Table 2 (RSE 6.9 percent)                      |
| `lvc` (V1 at 70 kg)       | log(26.0 L)          | Danielak 2017 Table 2 (RSE 14.0 percent)                     |
| `lq`  (Q)                 | log(2.25 L/h)        | Danielak 2017 Table 2 (RSE 22.2 percent; no WT covariate)    |
| `lvp` (V2 at 70 kg)       | log(9.93 L)          | Danielak 2017 Table 2 (RSE 9.0 percent)                      |
| `e_wt_cl`                 | fixed(0.75)          | Danielak 2017 Table 2 ("beta Cl, weight 0.75 (fixed)")       |
| `e_wt_vc`                 | fixed(1.0)           | Danielak 2017 Table 2 ("beta V1, weight 1 (fixed)")          |
| `e_wt_vp`                 | fixed(1.0)           | Danielak 2017 Table 2 ("beta V2, weight 1 (fixed)")          |
| `var(etalcl)`             | 0.065025             | Danielak 2017 Table 2 (omega_Cl 25.5 percent -> 0.255^2)     |
| `var(etalvc)`             | 0.264196             | Danielak 2017 Table 2 (omega_V1 51.4 percent -> 0.514^2)     |
| `cov(etalcl, etalvc)`     | 0.093596             | Danielak 2017 Table 2 (rho_Cl-V1 0.714 -> rho * omega1 * omega2) |
| `var(etalq)`              | 0.148996             | Danielak 2017 Table 2 (omega_Q 38.6 percent -> 0.386^2)      |
| `propSd`                  | 0.188                | Danielak 2017 Table 2 (proportional residual error)          |
| WT covariate equation     | `(WT/70)^beta`       | Danielak 2017 Eq. 9                                          |
| Two-compartment ODE       | linear IV-infusion   | Danielak 2017 Methods "Structural and error model selection" |

# Virtual cohort

Individual subject-level data are not publicly available. The virtual
cohort below mirrors the paper's Table 1 demographics: weights drawn from
a log-normal centered on the cohort mean (26.9 kg) and clipped to the
observed range (7.7-52 kg). Body surface area is derived from weight via
the Costeff (1966) pediatric formula `BSA = (4*WT + 7)/(WT + 90)`, which
gives BSA values within a few percent of the Mosteller method used by the
paper across the source weight range. The cohort is replicated across the
three dosing regimens of the original study (12 g/m^2 over 1 h; 12 g/m^2
over 2 h; 14 g/m^2 over 2 h), so per-regimen typical-value behaviour can
be compared.

```{r cohort}
set.seed(20260529)

n_per_arm <- 100L

# Weight distribution per Table 1 (mean 26.9 kg, SD 15.7 kg, range 7.7-52).
# Use a log-normal whose first two moments approximate the reported mean
# and SD on the natural scale, then truncate to the source range.
mu_wt <- 26.9
sd_wt <- 15.7
ln_sd  <- sqrt(log(1 + (sd_wt / mu_wt)^2))
ln_mu  <- log(mu_wt) - ln_sd^2 / 2

draw_weights <- function(n) {
  w <- numeric(0)
  while (length(w) < n) {
    candidate <- rlnorm(n, ln_mu, ln_sd)
    w <- c(w, candidate[candidate >= 7.7 & candidate <= 52])
  }
  w[seq_len(n)]
}

# Costeff pediatric BSA (m^2) from weight (kg). Within a few percent of
# Mosteller for the 8-52 kg pediatric range.
bsa_from_wt <- function(wt_kg) (4 * wt_kg + 7) / (wt_kg + 90)

regimens <- tibble::tibble(
  treatment      = c("12 g/m^2 in 1 h", "12 g/m^2 in 2 h", "14 g/m^2 in 2 h"),
  dose_g_per_m2  = c(12, 12, 14),
  infusion_h     = c( 1,  2,  2)
)

make_arm <- function(treatment, dose_g_per_m2, infusion_h,
                     n, id_offset) {
  wt <- draw_weights(n)
  bsa <- bsa_from_wt(wt)
  amt_mg <- dose_g_per_m2 * bsa * 1000      # g/m^2 * m^2 * (1000 mg / g)
  rate_mg_per_h <- amt_mg / infusion_h

  per_subject <- function(i) {
    ev <- rxode2::et(amt = amt_mg[i], cmt = "central",
                     rate = rate_mg_per_h[i]) |>
      rxode2::et(seq(0, 24, by = 0.1)) |>
      as.data.frame()
    ev$id        <- id_offset + i
    ev$WT        <- wt[i]
    ev$BSA       <- bsa[i]
    ev$treatment <- treatment
    ev
  }
  do.call(rbind, lapply(seq_len(n), per_subject))
}

events <- dplyr::bind_rows(
  make_arm(regimens$treatment[1], regimens$dose_g_per_m2[1],
           regimens$infusion_h[1], n_per_arm, id_offset = 0L),
  make_arm(regimens$treatment[2], regimens$dose_g_per_m2[2],
           regimens$infusion_h[2], n_per_arm, id_offset = 100L),
  make_arm(regimens$treatment[3], regimens$dose_g_per_m2[3],
           regimens$infusion_h[3], n_per_arm, id_offset = 200L)
)
stopifnot(!anyDuplicated(events[, c("id", "time", "evid")]))
```

# Simulation

Two simulations are run: a stochastic VPC-style population (full BSV) for
percentile plots, and a typical-value simulation (BSV zeroed) for the
median trajectory used in summary comparisons.

```{r simulate}
mod <- readModelDb("Danielak_2017_treosulfan")
mod_stoch   <- rxode2::rxode(mod)
mod_typical <- rxode2::zeroRe(rxode2::rxode(mod))

sim <- rxode2::rxSolve(
  mod_stoch,
  events = events,
  keep   = c("treatment", "WT", "BSA")
) |>
  as.data.frame()

sim_typical <- rxode2::rxSolve(
  mod_typical,
  events = events,
  keep   = c("treatment", "WT", "BSA")
) |>
  as.data.frame()
```

# Replicate Figure 2 -- individual concentration-time profiles

Danielak 2017 Figure 2 is a spaghetti plot of individual concentration-time
curves over the first 12 h, with treosulfan plasma concentration on a
log y-axis. The simulated stochastic cohort below reproduces the shape:
rapid rise during the 1-2 h infusion, then a multi-exponential decline
spanning roughly two orders of magnitude to 8-12 h post-dose.

```{r figure-2}
sim |>
  dplyr::filter(time > 0, time <= 12) |>
  dplyr::group_by(treatment, id) |>
  dplyr::sample_n(min(dplyr::n(), 60)) |>      # thin to ~60 points / subject
  dplyr::ungroup() |>
  ggplot(aes(time, Cc, group = id)) +
  geom_line(alpha = 0.20) +
  facet_wrap(~treatment) +
  scale_y_log10(limits = c(1, 1000)) +
  labs(x = "Time after start of infusion (h)",
       y = "Treosulfan plasma concentration (mg/L)",
       title = "Replicates Figure 2 of Danielak 2017 -- spaghetti plot of individual profiles",
       caption = "Stochastic simulation, 100 virtual subjects per dosing arm.")
```

# Replicate Figure 4 -- prediction-corrected VPC

Danielak 2017 Figure 4 shows a pcVPC across all subjects. The plot below
approximates the same form: simulated 5th / median / 95th percentiles of
the stochastic cohort by treatment arm.

```{r figure-4}
sim |>
  dplyr::filter(time > 0, time <= 12) |>
  dplyr::group_by(treatment, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~treatment) +
  scale_y_log10(limits = c(1, 1000)) +
  labs(x = "Time after start of infusion (h)",
       y = "Treosulfan plasma concentration (mg/L)",
       title = "VPC by dosing regimen",
       caption = "Replicates the spirit of Figure 4 (pcVPC) in Danielak 2017.")
```

# PKNCA validation

AUC0-inf, Cmax and Tmax are computed with PKNCA for each dosing arm,
grouped by treatment. The simulated AUCs are compared with literature
adult treosulfan AUC values for sanity checking.

```{r pknca}
sim_nca <- sim_typical |>
  dplyr::filter(!is.na(Cc), time >= 0) |>
  dplyr::select(id, time, Cc, treatment)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- PKNCA::pk.nca(nca_data)
nca_tbl  <- as.data.frame(nca_res$result)

nca_summary <- nca_tbl |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  dplyr::select(treatment, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    Cmax_mg_per_L    = median(cmax,       na.rm = TRUE),
    Tmax_h           = median(tmax,       na.rm = TRUE),
    AUCinf_mg_h_per_L = median(aucinf.obs, na.rm = TRUE),
    HalfLife_h       = median(half.life,  na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(nca_summary, digits = c(0, 1, 2, 1, 2),
             caption = "Median typical-value NCA parameters by dosing arm.")
```

## Comparison against published exposure

Danielak 2017 reports `CL = 14.7 L/h` at 70 kg with fixed allometric
exponent 0.75 on WT, and the source cohort mean weight is 26.9 kg, giving
a typical-CL of `14.7 * (26.9 / 70)^0.75 ~ 7.0 L/h`. For 12 g/m^2 at a
typical-cohort BSA of about 1.0 m^2, the daily dose is `12 g = 12000 mg`,
giving an expected single-dose `AUC0-inf = Dose / CL ~ 12000 / 7.0 ~ 1700
mg.h/L`. The 14 g/m^2 arm scales linearly to about `~2000 mg.h/L`. The
simulated AUC values above are consistent with these algebraic checks.
Scheulen et al. (cited by Danielak 2017) report dose-linear treosulfan
pharmacokinetics over the 20-56 g/m^2 range; the predicted AUC scaling
between the 12 g/m^2 and 14 g/m^2 arms (~14/12 = 1.17x) is also linear,
matching that prior literature.

# Assumptions and deviations

- The source paper used Mosteller-method BSA from observed height and
  weight for dose calculation; only weight is publicly available, so the
  virtual cohort uses the Costeff (1966) pediatric weight-only BSA formula
  `BSA = (4*WT + 7) / (WT + 90)`. The two formulas agree within a few
  percent across the source weight range (7.7-52 kg).
- The simulated cohort uses 100 subjects per dosing arm to give stable
  VPC percentiles; the source dataset has only 15 patients across all
  arms combined, so individual-level reproduction of the published
  Figure 3 (GOF plots) is not attempted.
- The IIV omegas in Danielak 2017 Table 2 are reported in percent. The
  paper documents an exponential / lognormal IIV model
  (`theta_ij = theta_j * exp(eta_ij)`), so the percent value is treated
  as `omega = 0.255` (the SD of eta on the log scale) and the internal
  variance is `omega^2` (not `log(CV^2 + 1)` -- that conversion applies
  when the paper reports the linear-space CV percent of `theta_ij`,
  whereas Monolix's IIV percent here is the SD of eta itself).
- The off-diagonal `omega_Cl-V1 = 71.4` is interpreted as a correlation
  coefficient (rho = 0.714) rather than a raw covariance; the resulting
  internal covariance is `rho * omega_Cl * omega_V1 = 0.0936`. This
  matches Monolix's default OMEGA-matrix output convention, in which the
  off-diagonal cells display correlations.
- The paper's IIV on V2 was estimated but removed from the final model
  because its standard error exceeded 100 percent (Danielak 2017 Results
  paragraph 1); V2 is therefore treated as a typical-value parameter
  here.
