---
title: "Remlifanserin (Darwish 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Remlifanserin (Darwish 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Darwish_2026_remlifanserin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Darwish_2026_remlifanserin"))$description`
* Article: <https://doi.org/10.1002/trc2.70254>
* Supplement (Supplemental Tables 1-6 and the NONMEM control stream):
  <https://www.ebi.ac.uk/europepmc/webservices/rest/PMC13133547/supplementaryFiles>

Remlifanserin (ACP-204) is a selective serotonin 2A (5-HT~2A~) receptor inverse
agonist in development for Alzheimer's disease psychosis and Lewy body dementia
psychosis. Darwish 2026 is the first population PK model for the compound.

## Units

The model is parameterised in **days**, exactly as Darwish 2026 Table 1 and the
Supplemental Methods control stream report it. The control stream's scaling
line reads `S2=V ;Dose amount = ug; Volume = L; Concentration = ng/mL`, so
**dose amounts are entered in micrograms**: the 60 mg target dose is `60000`.
Darwish 2026 Section 3.4 restates the identical estimates in hours (CL/F
35.4 L/h, K~a~ 0.692 1/h, ALAG1 0.720 h fasted); those are conversions of the
same numbers, not a separate parameterisation.

# Population

The model was fit to 3935 plasma concentrations from 209 healthy participants
pooled across seven Phase 1 studies of remlifanserin (Supplemental Table 1): a
single-ascending-dose study with a food arm (001), a multiple-ascending-dose
study that included an elderly cohort (002), a PET receptor-occupancy study
(003), itraconazole and carbamazepine drug-drug-interaction studies (004 and
005, of which only period 1 was retained), a crossover food-effect study (010),
and a `[14C]` mass-balance study (011). Oral doses spanned 10 to 180 mg, given
as single doses and as once-daily multiple doses for 10 days.

Participants were 19 to 75 years old (median 37), weighed 47 to 108 kg (median
77.6), and had a BMI of 18.5 to 31.9 kg/m^2^ (median 26.3). 30.6% (64/209) were
female. Race was 59.3% White, 33.0% Black or African American, 2.87% Asian and
4.78% Other. Renal and hepatic function were predominantly normal (eGFR 58.4 to
138 mL/min/1.73 m^2^, median 103). Only 18 participants were in the 65-75 year
older-adult stratum, which Darwish 2026 Section 4.3 flags as a limitation.
Baseline demographics are Supplemental Table 2; the fed/fasted record split
(71% fasted, 29% fed) is Supplemental Table 3.

The same information is available programmatically via
`readModelDb("Darwish_2026_remlifanserin")()$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Darwish_2026_remlifanserin.R`. The table
below collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Structural model: 1-compartment, first-order absorption, absorption lag, linear elimination | n/a | Section 3.1; Supplemental Methods `$SUBROUTINES ADVAN2 TRANS2` |
| `lcl` (CL/F) | `log(849)` L/day | Table 1, "CL/F (L/day)" = 849, RSE 2.84% |
| `lvc` (V/F, male) | `log(930)` L | Table 1, "V/F (L)" = 930, RSE 2.43% |
| `lka` (K~a~) | `log(16.6)` 1/day | Table 1, "K~a~ (1/day)" = 16.6, RSE 4.65% |
| `ltlag` (ALAG1, fasted) | `log(0.0300)` day | Table 1, "ALAG1 (day)" = 0.0300, RSE 1.91% |
| `e_sexf_vc` | -131 L | Table 1, "Additive shift in V/F for females" = -131, RSE 16.8% |
| `e_fed_tlag` | 0.290 | Table 1, "Proportional shift in ALAG1 for fed status" = 0.290, RSE 10.4% |
| `vc <- (exp(lvc) + e_sexf_vc * SEXF) * exp(etalvc)` | n/a | Section 3.2 equation `V/F_i = 930 L - 131 * female_i`; `$PK` block `TVV = TVVI + COV2`, `V = TVV*EXP(ETA(2))` |
| `tlag <- exp(ltlag) * (1 + e_fed_tlag * FED) * exp(etaltlag)` | n/a | Section 3.2 equation `ALAG1_i = 0.0300 d * (1 + 0.290 * fed_i)`; `$PK` block `TVALAG1 = THETA(4)*(1+THETA(5)*FED)` |
| `etalcl` variance | 0.1668 | Table 1, CL/F 42.6 %CV, back-transformed as `log(1 + 0.426^2)` |
| `etalvc` variance | 0.0981 | Table 1, V/F 32.1 %CV, back-transformed as `log(1 + 0.321^2)` |
| `etalcl`-`etalvc` covariance | 0.109 | Table 1, "Covariance (IIV in V/F, IIV in CL/F)" = 0.109, RSE 11.9% |
| `etalka` variance | 0.3754 | Table 1, K~a~ 67.5 %CV, back-transformed as `log(1 + 0.675^2)` |
| `etaltlag` variance | 0.01138 | Table 1, ALAG1 10.7 %CV, back-transformed as `log(1 + 0.107^2)` |
| `propSd` | 0.197 | Table 1, "Residual variability" = 0.0388 (a variance), reported as 19.7 %CV; `sqrt(0.0388)` = 0.1970 |

## Pinning the omega scale

Darwish 2026 prints interindividual variability only as a percent CV, so each
variance had to be back-transformed. Two independent checks establish that the
paper used the log-normal relation `omega^2 = log(1 + CV^2)` rather than the
`omega = CV` approximation.

```{r omega-scale}
cv <- c(cl = 0.426, vc = 0.321)            # Table 1, %CV column
covariance <- 0.109                         # Table 1, covariance row

omega2_lognormal <- log(1 + cv^2)           # candidate A
omega2_cvdirect  <- cv^2                    # candidate B

r_lognormal <- covariance / sqrt(prod(omega2_lognormal))
r_cvdirect  <- covariance / sqrt(prod(omega2_cvdirect))

# Table 1 footnote: "The calculated correlation coefficient (r) associated with
# covariance (IIV in V/F, IIV in CL/F) was 0.852 with r^2 = 0.726."
data.frame(
  scale = c("omega^2 = log(1 + CV^2)", "omega^2 = CV^2"),
  r     = c(r_lognormal, r_cvdirect),
  r2    = c(r_lognormal, r_cvdirect)^2,
  published_r = 0.852,
  published_r2 = 0.726
) |>
  knitr::kable(digits = 4, caption = "Which omega scale reproduces the Table 1 footnote?")

# The log-normal back-transform reproduces the published r to three decimals;
# the CV-squared alternative is off by more than 0.05. Gate on the footnote.
stopifnot(
  abs(r_lognormal - 0.852) < 0.002,
  abs(r_lognormal^2 - 0.726) < 0.002,
  abs(r_cvdirect - 0.852) > 0.02
)
```

The Supplemental Methods `$OMEGA` initial estimates (0.16 for CL/F, 0.09 for
V/F, 0.10 for the covariance, 0.30 for K~a~, 0.01 for ALAG1) sit next to the
back-transformed final values (0.1668, 0.0981, 0.109, 0.3754, 0.01138),
independently confirming that these are NONMEM variances on the log scale.

# Typical-value replication

Darwish 2026 Section 3.4 restates the final estimates for a typical
participant. Because these are deterministic functions of the `ini()` block,
they can be gated tightly.

```{r typical-values}
mod <- readModelDb("Darwish_2026_remlifanserin")
mod_typical <- rxode2::zeroRe(mod)

# 60 mg = 60000 ug single oral dose, dense grid for Cmax / Tmax resolution.
typical_events <- tidyr::expand_grid(
  arm  = c("Male, fasted", "Male, fed", "Female, fasted", "Female, fed"),
  time = c(0, seq(0, 25, by = 0.0005))
) |>
  dplyr::distinct(arm, time) |>
  dplyr::mutate(
    id   = as.integer(factor(arm)),
    SEXF = as.integer(grepl("Female", arm)),
    FED  = as.integer(grepl("fed$", arm)),
    evid = 0L, amt = NA_real_, cmt = "central"
  ) |>
  dplyr::bind_rows(
    tibble::tibble(
      arm  = c("Male, fasted", "Male, fed", "Female, fasted", "Female, fed")
    ) |>
      dplyr::mutate(
        id   = as.integer(factor(arm)),
        SEXF = as.integer(grepl("Female", arm)),
        FED  = as.integer(grepl("fed$", arm)),
        time = 0, evid = 1L, amt = 60000, cmt = "depot"
      )
  ) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim_typical <- rxode2::rxSolve(
  mod_typical, events = typical_events, keep = "arm", returnType = "data.frame"
)

typical_summary <- sim_typical |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    `CL/F (L/h)`  = dplyr::first(cl) / 24,
    `V/F (L)`     = dplyr::first(vc),
    `Ka (1/h)`    = dplyr::first(ka) / 24,
    `ALAG1 (h)`   = dplyr::first(tlag) * 24,
    `t1/2 (h)`    = log(2) / dplyr::first(kel) * 24,
    `Cmax (ng/mL)` = max(Cc),
    `Tmax (h)`    = time[which.max(Cc)] * 24,
    .groups = "drop"
  )

knitr::kable(
  typical_summary, digits = 3,
  caption = "Typical-value predictions for a 60 mg single oral dose."
)
```

```{r typical-gate}
male_fasted <- typical_summary[typical_summary$arm == "Male, fasted", ]
fem_fasted  <- typical_summary[typical_summary$arm == "Female, fasted", ]
male_fed    <- typical_summary[typical_summary$arm == "Male, fed", ]

# Darwish 2026 Section 3.4, verbatim: "The estimated population mean Ka of
# 0.692/h ... The estimated population mean for ALAG1 is 0.720 h under the
# fasted state and is predicted to increase to 0.929 h with food intake. CL/F
# for a typical participant is estimated to be 35.4 L/h. V/F is predicted to be
# 930 L for a typical male participant and 799 L for a typical female
# participant."
stopifnot(
  abs(male_fasted$`CL/F (L/h)` - 35.4)   < 0.05,
  abs(male_fasted$`Ka (1/h)`   - 0.692)  < 0.001,
  abs(male_fasted$`ALAG1 (h)`  - 0.720)  < 0.001,
  abs(male_fed$`ALAG1 (h)`     - 0.929)  < 0.001,
  abs(male_fasted$`V/F (L)`    - 930)    < 0.5,
  abs(fem_fasted$`V/F (L)`     - 799)    < 0.5
)

# Section 4.1: food delays Tmax by "~13 min". The lag shift is the only
# mechanism, so the Tmax shift equals the ALAG1 shift exactly.
tmax_shift_min <- (male_fed$`Tmax (h)` - male_fasted$`Tmax (h)`) * 60
cat(sprintf("Food-induced Tmax shift: %.1f min (Darwish 2026 Section 4.1: ~13 min)\n",
            tmax_shift_min))
stopifnot(abs(tmax_shift_min - 13) < 1.5)
```

## Mass balance

Only oral data were available, so `cl` and `vc` are apparent values and the
dose enters the depot unscaled (Section 2.3). The identity
`CL/F * AUC(0-Inf) = dose` must therefore hold exactly. This is the gate that
would catch an `alag()` that silently zeroed the depot, or an unnoticed
`linCmt()` auto-solve that discarded the explicit ODEs.

```{r mass-balance}
stopifnot(is.null(rxode2::rxode(mod)$linCmt))  # explicit ODEs, not auto-solved

mass_balance <- sim_typical |>
  dplyr::group_by(arm) |>
  dplyr::arrange(time, .by_group = TRUE) |>
  dplyr::summarise(
    # trapezoidal AUC over the observed grid plus the analytic terminal tail
    auc = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2) +
      dplyr::last(Cc) / dplyr::first(kel),
    ratio = dplyr::first(cl) * auc / 60000,
    .groups = "drop"
  )

knitr::kable(mass_balance, digits = 6,
             caption = "CL/F * AUC(0-Inf) / dose; must be 1.")
stopifnot(max(abs(mass_balance$ratio - 1)) < 1e-3)
```

## Time to steady state

Section 4.2 states that "At daily doses of 60 mg of remlifanserin, participants
achieved steady-state exposure levels within 6 days."

```{r time-to-ss}
ss_events <- dplyr::bind_rows(
  tibble::tibble(time = 0:20, evid = 1L, amt = 60000, cmt = "depot"),
  tibble::tibble(time = seq(0, 21, by = 0.01), evid = 0L,
                 amt = NA_real_, cmt = "central")
) |>
  dplyr::mutate(id = 1L, SEXF = 0L, FED = 0L) |>
  dplyr::arrange(time, dplyr::desc(evid))

# rxSolve returns observation rows only and does not carry `evid` through to
# its output, so there is nothing to filter here.
ss <- rxode2::rxSolve(mod_typical, events = ss_events, returnType = "data.frame")

by_day <- ss |>
  dplyr::mutate(day = ceiling(pmax(time, 1e-9))) |>
  dplyr::filter(day >= 1, day <= 20) |>
  dplyr::group_by(day) |>
  dplyr::summarise(cmax = max(Cc), auc_tau = mean(Cc), .groups = "drop")

plateau <- by_day$auc_tau[by_day$day == 20]
first_day_within <- function(pct) min(by_day$day[by_day$auc_tau / plateau >= pct])

cat(sprintf("AUC(0-24) reaches 90%% of plateau on day %d, 95%% on day %d, 99%% on day %d\n",
            first_day_within(0.90), first_day_within(0.95), first_day_within(0.99)))
stopifnot(first_day_within(0.95) <= 6, first_day_within(0.99) <= 6)

ggplot(by_day, aes(day, auc_tau / plateau)) +
  geom_line() + geom_point() +
  geom_hline(yintercept = 1, linetype = "solid", colour = "grey40") +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = 6, linetype = "dotted", colour = "firebrick") +
  labs(x = "Day of once-daily 60 mg dosing", y = "AUC(0-24) as a fraction of plateau",
       caption = "Dotted red line: the 6 days quoted in Darwish 2026 Section 4.2.")
```

## Dose proportionality (Figure 3)

Section 2.7 fits the power model `ln(Y) = intercept + beta * ln(dose)` and
declares proportionality when the 90% CI of `beta` lies within
`(1 + ln(0.8)/ln(r), 1 + ln(1.25)/ln(r))`, with `r` the ratio of the highest to
the lowest dose. The packaged model is structurally linear, so this reproduces
the paper's conclusion by construction rather than as an independent test; it
is included to confirm that no dose-dependent term crept into the encoding.

```{r dose-proportionality}
doses_mg <- c(10, 20, 40, 60, 90, 120, 130, 180)

dp_events <- tidyr::expand_grid(dose_mg = doses_mg, time = seq(0, 25, by = 0.005)) |>
  dplyr::mutate(id = as.integer(factor(dose_mg)), evid = 0L,
                amt = NA_real_, cmt = "central") |>
  dplyr::bind_rows(
    tibble::tibble(dose_mg = doses_mg) |>
      dplyr::mutate(id = as.integer(factor(dose_mg)), time = 0, evid = 1L,
                    amt = dose_mg * 1000, cmt = "depot")
  ) |>
  dplyr::mutate(SEXF = 0L, FED = 0L) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

dp <- rxode2::rxSolve(mod_typical, events = dp_events, keep = "dose_mg",
                      returnType = "data.frame") |>
  dplyr::group_by(dose_mg) |>
  dplyr::filter(time <= 1) |>
  dplyr::summarise(
    cmax    = max(Cc),
    auc0_24 = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    .groups = "drop"
  )

r_ratio <- max(doses_mg) / min(doses_mg)
lo <- 1 + log(0.8) / log(r_ratio)
hi <- 1 + log(1.25) / log(r_ratio)

beta <- vapply(c(cmax = "cmax", auc0_24 = "auc0_24"), function(y) {
  unname(coef(lm(log(dp[[y]]) ~ log(dp$dose_mg)))[2])
}, numeric(1))

data.frame(
  parameter = c("Cmax", "AUC0-24"),
  beta = beta,
  criterion_low = lo,
  criterion_high = hi,
  within = beta > lo & beta < hi
) |>
  knitr::kable(digits = 4, caption = "Dose-proportionality exponents (Figure 3 criterion).")

stopifnot(all(abs(beta - 1) < 1e-3), all(beta > lo & beta < hi))
```

# Virtual cohort

Original observed data are not publicly available. Two virtual cohorts are
used, each of 200 participants per arm.

* **Analysis-like cohort** -- 200 participants with the sex balance (30.6%
  female) and fed-record fraction (29%) of the pooled analysis dataset
  (Supplemental Tables 2 and 3). Used for the post hoc parameter comparison
  against Table 2 and for the NCA validation.
* **Simulation cohort** -- the 2 x 2 factorial of sex and fed status at 200 per
  cell, mirroring the 1:1 random assignment Darwish 2026 Section 2.6 used for
  its virtual population of 1000 (500 per group for each categorical
  covariate). Used for the steady-state exposure comparison.

```{r cohort}
# set.seed() seeds R's RNG, not rxode2's simulation RNG, and rxode2's streams
# are partitioned per solver thread -- so this cohort is reproducible on this
# machine and different on a machine with a different thread count. Every
# assertion below is written to hold for any cohort the model can produce.
set.seed(20260913)
rxode2::rxSetSeed(20260913)

expand_events <- function(subjects, dose_ug, dose_times, obs_times) {
  dosing <- tidyr::expand_grid(subjects, time = dose_times) |>
    dplyr::mutate(evid = 1L, amt = dose_ug, cmt = "depot")
  obs <- tidyr::expand_grid(subjects, time = obs_times) |>
    dplyr::mutate(evid = 0L, amt = NA_real_, cmt = "central")
  dplyr::bind_rows(dosing, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

n_arm <- 200L

# Analysis-like cohort: single 60 mg dose, followed to day 5.
subj_analysis <- tibble::tibble(
  id   = seq_len(n_arm),
  SEXF = as.integer(seq_len(n_arm) <= round(0.306 * n_arm)),
  FED  = as.integer(seq_len(n_arm) %% 100L < 29L),
  arm  = "Analysis-like"
)
ev_analysis <- expand_events(subj_analysis, 60000, 0, seq(0, 5, by = 0.02))

# Simulation cohort: 2 x 2 factorial, 60 mg once daily for 14 days.
subj_sim <- tidyr::expand_grid(SEXF = 0:1, FED = 0:1, k = seq_len(n_arm)) |>
  dplyr::mutate(
    id  = dplyr::row_number(),
    arm = paste0(ifelse(SEXF == 1, "Female", "Male"), ", ",
                 ifelse(FED == 1, "fed", "fasted"))
  ) |>
  dplyr::select(id, SEXF, FED, arm)

ev_sim <- expand_events(
  subj_sim, 60000,
  dose_times = 0:13,
  obs_times  = sort(unique(c(seq(0, 13, by = 0.25), seq(13, 14, by = 0.005))))
)

stopifnot(
  !anyDuplicated(unique(ev_analysis[, c("id", "time", "evid")])),
  !anyDuplicated(unique(ev_sim[, c("id", "time", "evid")])),
  nrow(dplyr::count(subj_sim, arm)) == 4L,
  all(dplyr::count(subj_sim, arm)$n == n_arm)
)
```

```{r simulate}
sim_analysis <- rxode2::rxSolve(mod, events = ev_analysis, keep = c("arm", "SEXF", "FED"),
                                returnType = "data.frame")
sim_ss <- rxode2::rxSolve(mod, events = ev_sim, keep = c("arm", "SEXF", "FED"),
                          returnType = "data.frame")
stopifnot(!anyNA(sim_analysis$Cc), !anyNA(sim_ss$Cc), all(sim_ss$Cc >= 0))
```

# Replicate published figures

## Figure 2 -- concentration-time profiles stratified by food status and by sex

Figure 2 of Darwish 2026 is a prediction-corrected VPC of the pooled dataset
stratified by fasted (A) versus fed (B) and by male (C) versus female (D). The
panels below show the corresponding simulated median and 5th/95th percentiles
over the first two days of once-daily 60 mg dosing.

```{r figure-2}
sim_ss |>
  dplyr::filter(time <= 2) |>
  tidyr::pivot_longer(c(FED, SEXF), names_to = "stratum_var", values_to = "level") |>
  dplyr::mutate(
    stratum = dplyr::case_when(
      stratum_var == "FED"  & level == 0 ~ "A. Fasted",
      stratum_var == "FED"  & level == 1 ~ "B. Fed",
      stratum_var == "SEXF" & level == 0 ~ "C. Male",
      TRUE                               ~ "D. Female"
    )
  ) |>
  dplyr::group_by(stratum, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time * 24, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue4") +
  facet_wrap(~stratum) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Remlifanserin concentration (ng/mL)",
       caption = "Replicates the stratification of Figure 2 of Darwish 2026 (60 mg once daily).")
```

## Table 2 -- post hoc pharmacokinetic parameters

Table 2 summarises the post hoc PK parameters of the 209 analysed
participants. Because the model's random effects are log-normal, the expected
values of these summaries are available in closed form, which makes this a
deterministic comparison independent of any simulated draw. The analysis-like
cohort's empirical values are shown alongside for context but are not gated.

```{r table-2-analytic}
om <- c(cl = 0.1668, vc = 0.0981, ka = 0.3754, tlag = 0.01138)
cov_cl_vc <- 0.109
p_female  <- 0.306    # Supplemental Table 2
p_fed     <- 0.29     # Supplemental Table 3, record-weighted

e_vc_typ  <- 930 - 131 * p_female         # E[typical V/F] over the sex mixture
e_inv_vc  <- (1 - p_female) / 930 + p_female / 799
# etalcl and etalvc are correlated, so E[exp(etalcl - etalvc)] uses their
# combined variance: var(cl) + var(vc) - 2 * cov.
var_diff  <- om[["cl"]] + om[["vc"]] - 2 * cov_cl_vc

analytic <- c(
  `CL/F (L/day)` = 849 * exp(om[["cl"]] / 2),
  `V/F (L)`      = e_vc_typ * exp(om[["vc"]] / 2),
  `Kel (1/day)`  = 849 * e_inv_vc * exp(var_diff / 2),
  `T1/2 (h)`     = log(2) * 24 * (e_vc_typ / 849) * exp(var_diff / 2),
  `Ka (1/day)`   = 16.6 * exp(om[["ka"]] / 2),
  `ALAG1 (day)`  = 0.0300 * (1 + 0.290 * p_fed) * exp(om[["tlag"]] / 2)
)

published_mean <- c(`CL/F (L/day)` = 919.85, `V/F (L)` = 931.83,
                    `Kel (1/day)` = 0.98, `T1/2 (h)` = 17.75,
                    `Ka (1/day)` = 19.61, `ALAG1 (day)` = 0.0332)

per_subject <- sim_analysis |>
  dplyr::distinct(id, .keep_all = TRUE) |>
  dplyr::transmute(cl, vc, kel, t_half = log(2) / kel * 24, ka, tlag)

empirical_mean <- c(
  `CL/F (L/day)` = mean(per_subject$cl), `V/F (L)` = mean(per_subject$vc),
  `Kel (1/day)` = mean(per_subject$kel), `T1/2 (h)` = mean(per_subject$t_half),
  `Ka (1/day)` = mean(per_subject$ka),   `ALAG1 (day)` = mean(per_subject$tlag)
)

t2 <- data.frame(
  Parameter = names(published_mean),
  `Published mean (Table 2)` = unname(published_mean),
  `Model expected mean` = unname(analytic[names(published_mean)]),
  `Cohort mean (n = 200)` = unname(empirical_mean[names(published_mean)]),
  `% diff (analytic)` = unname(100 * (analytic[names(published_mean)] / published_mean - 1)),
  check.names = FALSE
)
knitr::kable(t2, digits = c(0, 4, 4, 4, 2),
             caption = "Table 2 post hoc means vs. the model's closed-form expectations.")

# Structural gate: a mis-transcribed theta or a mis-scaled omega moves these by
# tens of percent. The closed-form column is deterministic, so a tight bound is
# appropriate here (it is not a random-cohort extreme).
stopifnot(max(abs(t2$`% diff (analytic)`)) < 5)
```

Every closed-form expectation lands within a few percent of the published post
hoc mean, which jointly confirms the thetas, the covariate coefficients, the
omega back-transform and the CL/V covariance. `Kel` and `T1/2` are the sharpest
of these: both depend on the CL/V *covariance* through
`exp((var_cl + var_vc - 2 * cov) / 2)`, and would be visibly wrong if the
covariance had been dropped or its sign flipped.

## Figure 4 -- clinical relevance of the covariate effects

Figure 4 presents geometric mean ratios with 90% CIs for C~max,ss~ and
AUC~0-24,ss~ against clinical-relevance bounds of 0.8 and 1.25. The paper's
conclusion is that "neither fed/fasted status nor sex exhibits a clinically
relevant effect on steady-state remlifanserin exposures".

```{r figure-4}
tau_window <- function(d, lo = 13, hi = 14) dplyr::filter(d, time >= lo, time <= hi)

exposure <- sim_ss |>
  tau_window() |>
  dplyr::group_by(id, arm, SEXF, FED) |>
  dplyr::arrange(time, .by_group = TRUE) |>
  dplyr::summarise(
    cmax_ss = max(Cc),
    # ng/mL * day -> ng*h/mL
    auc_ss  = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2) * 24,
    .groups = "drop"
  )

gmr <- function(num, den) {
  lr <- log(num) - mean(log(den))
  n1 <- length(num); n2 <- length(den)
  s  <- sqrt(var(log(num)) / n1 + var(log(den)) / n2)
  m  <- mean(log(num)) - mean(log(den))
  c(gmr = exp(m), lower = exp(m - 1.645 * s), upper = exp(m + 1.645 * s))
}

contrasts <- dplyr::bind_rows(
  data.frame(metric = "Cmax,ss", contrast = "Fed / fasted",
             t(gmr(exposure$cmax_ss[exposure$FED == 1], exposure$cmax_ss[exposure$FED == 0]))),
  data.frame(metric = "AUC0-24,ss", contrast = "Fed / fasted",
             t(gmr(exposure$auc_ss[exposure$FED == 1], exposure$auc_ss[exposure$FED == 0]))),
  data.frame(metric = "Cmax,ss", contrast = "Female / male",
             t(gmr(exposure$cmax_ss[exposure$SEXF == 1], exposure$cmax_ss[exposure$SEXF == 0]))),
  data.frame(metric = "AUC0-24,ss", contrast = "Female / male",
             t(gmr(exposure$auc_ss[exposure$SEXF == 1], exposure$auc_ss[exposure$SEXF == 0])))
)

knitr::kable(contrasts, digits = 3,
             caption = "Geometric mean ratios with 90% CIs (Figure 4 of Darwish 2026).")

ggplot(contrasts, aes(gmr, contrast)) +
  geom_vline(xintercept = 1, linetype = "solid", colour = "grey40") +
  geom_vline(xintercept = 0.8, linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = 1.25, linetype = "dashed", colour = "grey40") +
  geom_pointrange(aes(xmin = lower, xmax = upper)) +
  facet_wrap(~metric) +
  labs(x = "Geometric mean ratio (90% CI)", y = NULL,
       caption = "Replicates Figure 4 of Darwish 2026; dashed lines are the 0.8-1.25 bounds.")
```

```{r figure-4-gate}
# The paper's own conclusion: every point estimate and 90% CI inside 0.8-1.25.
stopifnot(all(contrasts$lower > 0.8), all(contrasts$upper < 1.25))
```

The two mechanisms behind these ratios are worth stating explicitly, because
they are what makes the ratios insensitive to which cohort is drawn.
AUC~0-24,ss~ is `dose / (CL/F)` for a linear model at steady state, and neither
covariate acts on CL/F, so both AUC ratios are 1 up to sampling noise. Food
acts only through the absorption lag, which time-shifts a periodic steady state
without changing its peak, so the fed/fasted C~max,ss~ ratio is also 1. The one
ratio that is genuinely non-unity is female/male C~max,ss~, driven by the
`-131 L` volume shift; it is the closest to a bound and still comfortably
inside it.

# PKNCA validation

NCA is run on the analysis-like single-dose cohort using `Cc` (the individual
prediction, which carries no residual error) so that the terminal slope is not
corrupted by simulated assay noise.

```{r pknca}
sim_nca <- sim_analysis |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

# Guarantee a time = 0 row per subject; pre-dose Cc = 0 is correct for an
# extravascular dose.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

dose_df <- ev_analysis |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, arm)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "aucinf.obs", "half.life")) |>
  dplyr::group_by(PPTESTCD) |>
  dplyr::summarise(
    median = median(PPORRES, na.rm = TRUE),
    p10 = quantile(PPORRES, 0.10, na.rm = TRUE),
    p90 = quantile(PPORRES, 0.90, na.rm = TRUE),
    .groups = "drop"
  )
knitr::kable(nca_wide, digits = 4,
             caption = "PKNCA results, 60 mg single oral dose (time in days).")

# Internal consistency: NCA AUCinf must recover dose / (CL/F) subject by
# subject. This is the strongest available check on the whole solve-and-NCA
# pipeline and holds exactly, not statistically.
auc_by_id <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD == "aucinf.obs") |>
  dplyr::select(id, aucinf = PPORRES) |>
  dplyr::left_join(dplyr::distinct(sim_analysis, id, cl), by = "id") |>
  dplyr::mutate(ratio = cl * aucinf / 60000)
cat(sprintf("CL/F * AUCinf / dose: median %.4f, range %.4f-%.4f\n",
            median(auc_by_id$ratio), min(auc_by_id$ratio), max(auc_by_id$ratio)))
stopifnot(abs(median(auc_by_id$ratio) - 1) < 0.01,
          quantile(abs(auc_by_id$ratio - 1), 0.95) < 0.02)
```

## Comparison against published NCA

Darwish 2026 reports no NCA C~max~ or AUC values for the pooled dataset -- its
Table 2 is a summary of post hoc *model* parameters. The one parameter with a
direct NCA analogue is the terminal half-life, whose published post hoc median
is 17.45 h.

```{r nca-compare}
published_nca <- data.frame(arm = "Analysis-like", half.life = 17.45)

nca_hours <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD == "half.life") |>
  dplyr::mutate(PPORRES = PPORRES * 24)   # days -> hours

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_hours,
  reference = published_nca,
  by = "arm",
  units = c(half.life = "h"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Simulated vs. published half-life. * differs by >20%.")

sim_half_life <- median(nca_hours$PPORRES, na.rm = TRUE)
cat(sprintf("Simulated median t1/2 %.2f h vs. published post hoc median 17.45 h (%.1f%%)\n",
            sim_half_life, 100 * (sim_half_life / 17.45 - 1)))
stopifnot(abs(sim_half_life / 17.45 - 1) < 0.10)
```

# Assumptions and deviations

* **Interindividual variances were back-transformed, not transcribed.**
  Darwish 2026 Table 1 prints IIV only as a percent CV. Each `ini()` variance
  is `log(1 + CV^2)`. The choice of scale is not an assumption: the "Pinning
  the omega scale" section above shows it is the only one of the two candidates
  that reproduces the Table 1 footnote's *r* = 0.852 / *r*^2^ = 0.726, and the
  Supplemental Methods `$OMEGA` initial estimates corroborate it independently.
* **Values come from Table 1, not from the supplement's `$THETA` block.** The
  Supplemental Methods control stream is the initial-estimate version (THETA
  845, 965, 16.7, 0.03, 0.289, -222; `$SIGMA` 0.0389). Every packaged value is
  the final estimate from Table 1. The control stream was used only for the
  model *structure* -- the additive-inside-the-eta placement of the sex effect
  on V/F, the proportional food effect on ALAG1, `S2 = V`, and the purely
  proportional `$ERROR` block.
* **Dose amounts are in micrograms.** This follows the control stream's
  `S2=V ;Dose amount = ug; Volume = L; Concentration = ng/mL` comment. A 60 mg
  dose is entered as `60000`. Entering `60` instead would under-predict
  concentrations 1000-fold.
* **Covariates screened but not retained are recorded, not encoded.** Age,
  weight, BMI, BSA, race, eGFR, renal-function category, hepatic-function
  category, ALT, AST and total bilirubin were all screened by Darwish 2026
  Section 2.4 and dropped. They appear in the model file's
  `covariatesDataExcluded` list so the provenance of the covariate search is
  preserved without declaring unused covariates. The supplement's `$PK` block
  still computes `RACEB`, `GFRCKDEP2` and `RFCAT2` as leftovers of the stepwise
  search; none enters a parameter equation, so none is encoded here.
* **Fed status is a per-record covariate.** Table 2's footnote states that
  ALAG1 is non-stationary per participant and that some participants in the
  food-effect study had more than one distinct ALAG1 value. The virtual cohorts
  here assign one fed status per subject, which is adequate for the covariate
  contrasts but does not reproduce the within-subject crossover of study 010.
  The record-weighted fed fraction of 29% (Supplemental Table 3) was used for
  the closed-form ALAG1 expectation; Table 2's own ALAG1 mean implies a
  fed fraction nearer 35% across its 242 records, which is the main reason the
  ALAG1 row of the Table 2 comparison is the least tight of the six.
* **Below-quantitation handling is not replicated.** Darwish 2026 excluded the
  1.3% of samples below the 0.10 ng/mL LLOQ as missing. Simulated `Cc` values
  carry no assay noise and are never censored here, so the NCA terminal slope
  is estimated from a noiseless profile. This makes the half-life comparison a
  check on the model's `CL/V` ratio rather than on the paper's BLQ policy.
* **Post hoc estimates shrink; simulated parameters do not.** The Table 2
  comparison is therefore made on means and medians, where shrinkage has
  little effect, and not on standard deviations or on the min/max columns,
  where it has a large one.
* **The dose-proportionality check is structural.** A one-compartment linear
  model cannot produce a non-unity power exponent, so reproducing Figure 3's
  conclusion confirms only that no dose-dependent term was introduced during
  encoding. It is not independent evidence for the paper's finding.
* **No parameter was tuned.** Every value in `ini()` is a verbatim Table 1
  estimate or a documented back-transform of one.
