---
title: "Trofinetide, DAFFODIL pediatric 2-4 years (Darwish 2025b)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trofinetide, DAFFODIL pediatric 2-4 years (Darwish 2025b)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)

# Parse the packaged models once and reuse them for every metadata reference
# below. `readModelDb()` returns the raw model *function*; calling it directly
# (e.g. `readModelDb("X")()$population`) fails because `ini({})` / `model({})`
# are not defined outside an rxode2 parsing context.
mod_meta  <- rxode2::rxode(readModelDb("Darwish_2025b_trofinetide"))
prev_meta <- rxode2::rxode(readModelDb("Darwish_2025a_trofinetide"))
```

# Model and source

* Citation: `r mod_meta$reference`
* Description: `r mod_meta$description`
* Article: [Adv Ther. 2025;42(2):1009-1025](https://doi.org/10.1007/s12325-024-03058-7)

Trofinetide is the first approved treatment for Rett syndrome (RTT), dosed
orally twice daily on a body-weight-banded schedule. Acadia and Simulations Plus
built a 13-study population pharmacokinetic model to support the weight bands
used in the phase 3 LAVENDER study (girls and women aged 5-20 years); that model
is packaged separately as `Darwish_2025a_trofinetide`.

This vignette covers the **update**: the same model re-estimated after adding
114 whole-blood concentrations from 13 girls aged 2-4 years enrolled in the
phase 2/3 DAFFODIL study, used to confirm that the DAFFODIL weight bands
(5 g BID for `>=` 9 to `<` 12 kg, 6 g BID for `>=` 12 to `<` 20 kg) put
steady-state exposure inside the 800-1200 ug*h/mL target window.

## A transcription hazard specific to this paper

The Table 2 footnote of Darwish 2025b prints a block headed "Model equations:"
that gives typical-value expressions for F1, ka, CL, Vc, and Vp. **Those
equations carry the *earlier* model's coefficients, not the ones in the table
they annotate.** The footnote reads `F1 = 0.828 ...`, `ka = 0.391 ...`,
`CL = 11.8 x (WTKG/58)^0.443 ...`, `Vc = 24.9 ...`, `Vp = 35.3 ...`, all of
which are the Darwish 2025a values; the "Population mean estimate" column beside
them reads 0.832, 0.394, 11.7, 0.486, 25.0, and 35.4.

The table wins, and the paper's own prose proves it -- see the cross-model gate
below, which reproduces the two published change figures exactly. The model file
takes only the *functional form* from the footnote and every *coefficient* from
the table.

# Population

The pooled analysis dataset comprised 5709 trofinetide whole-blood
concentrations from 455 participants across 14 clinical studies: eight phase 1
studies, four phase 2 studies in RTT, fragile X syndrome (FXS) and traumatic
brain injury (TBI), the phase 3 LAVENDER study, and the phase 2/3 DAFFODIL
study. The cohort was 156 healthy volunteers, 198 patients with RTT (female
only), 57 with TBI (male only), and 44 with FXS (male only), predominantly
female (56.9%), with a mean age of 21.8 years (range 2-64) and mean body weight
of 56.5 kg (range 9.8-140).

The DAFFODIL subgroup that motivated the update was 13 girls (92.3% white,
7.7% Asian) with a mean age of 3 years (range 2-4) and mean baseline body weight
of 13.4 kg (range 9.8-18.1). Concentrations were measured by LC-MS/MS in
lithium-heparinized whole blood, LLOQ 0.100 ug/mL and ULOQ 100 ug/mL.

```{r population}
str(mod_meta$population)
```

# Source trace

Every value below is transcribed from the "Population mean estimate" column of
Darwish 2025b Table 2. The reference values 58 kg, 22.4 years, and
124 mL/min/1.73 m^2 are the analysis-population medians per the Table 2
footnote. The *form* of each covariate term -- a proportional shift
`(1 + theta * indicator)` and power terms on WT, GFR and AGE -- comes from the
footnote equations; their coefficients do not (see above).

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL) | 11.7 L/h | Table 2, "CL (L/h)" |
| `e_wt_cl` | 0.486 | Table 2, "Covariate exponent of weight on CL" |
| `e_crcl_cl` | 0.272 | Table 2, "Covariate exponent of GFR on CL" |
| `e_rett_cl` | -0.146 | Table 2, "Shift in CL for RTT = 1" |
| `e_tbi_cl` | 0.229 | Table 2, "Shift in CL for TBI = 1" |
| `lvc` (Vc) | 25.0 L | Table 2, "Vc (L)" |
| `e_age_vc` | 0.556 | Table 2, "Covariate exponent of age on Vc" |
| `e_fxs_vc` | 1.16 | Table 2, "Shift in Vc for FXS = 1" |
| `lq` (Q) | 1.42 L/h | Table 2, "Q (L/h)" |
| `lvp` (Vp) | 35.4 L | Table 2, "Vp (L)" |
| `e_rett_vp` | 0.805 | Table 2, "Shift in Vp for RTT = 1" |
| `e_tbi_vp` | -0.753 | Table 2, "Shift in Vp for TBI = 1" |
| `lka` (ka) | 0.394 1/h | Table 2, "ka (1/h)" |
| `e_fed_ka` | -0.0969 | Table 2, "Shift in ka for FED" |
| `lfdepot` (F1) | 0.832 | Table 2, "F1" |
| `e_fed_f` | -0.133 | Table 2, "Shift in F1 for FED" |
| `e_dose18g_f` | -0.132 | Table 2, "Shift in F1 for 18 g dose" |
| `e_dose24g_f` | -0.284 | Table 2, "Shift in F1 for 24 g dose" |
| `e_diarrhea_f` | -0.157 | Table 2, "Shift in F1 for diarrhea" |
| `etalcl` | 0.0182 | Table 2, "Interindividual variability / CL"; printed as 13.6 %CV |
| `etalvc` | 0.0906 | Table 2, "Interindividual variability / Vc"; printed as 30.8 %CV |
| `etalq` | 0.355 | Table 2, "Interindividual variability / Q"; printed as 65.3 %CV |
| `etalvp` | 0.0874 | Table 2, "Interindividual variability / Vp"; printed as 30.2 %CV |
| `etalfdepot` | 0.04 | Table 2, "Interindividual variability / F1"; printed as 20.2 %CV |
| `expSdHealthy` | sqrt(0.0788) = 0.281 | Table 2, "Residual variability / Healthy subjects"; printed as 28.1 %CV |
| `expSdDisease` | sqrt(0.140) = 0.374 | Table 2, "Residual variability / Patients with RTT, TBI or FXS"; printed as 37.4 %CV |
| Structural model (2-cmt, first-order absorption, linear elimination) | n/a | Methods, "Population Pharmacokinetic Model Development" |
| Proportional-shift form `(1 + theta * indicator)` | n/a | Table 2 footnote equations (form only) |
| Exponential IIV, `%CV = sqrt(exp(omega^2) - 1) * 100` | n/a | Methods Eq. (1) |
| Log/exponential residual error | n/a | Methods Eq. (2) |

Two transcription points are worth spelling out, because Table 2 reports each
variability quantity twice and the two columns are on different scales.

**The interindividual-variability block holds variances.** The block near the
bottom of Table 2 lists 0.0182, 0.0906, 0.355, 0.0874, and 0.04 with no units.
Methods Eq. (1) defines `%CV = sqrt(exp(omega^2) - 1) * 100`; applying it to
those five numbers returns 13.6, 30.8, 65.3, 30.2, and 20.2 %CV, which is
exactly the "Variability / Estimate" column beside each structural parameter.
They are therefore the NONMEM `$OMEGA` variances and are used directly.

**The residual-variability rows hold variances too, but with a different %CV
convention.** The tabulated 0.0788 and 0.140 map to the printed 28.1 %CV and
37.4 %CV through the plain `sqrt()`, not through Eq. (1)
(`sqrt(0.0788) = 0.281`, `sqrt(0.140) = 0.374`). `lnorm()` takes a log-scale
SD, so the model file enters `sqrt(0.0788)` and `sqrt(0.140)`.

```{r verify-variability}
# Reproduce both %CV conventions from the tabulated variances, confirming which
# column is which. Deterministic; no simulation involved.
iiv_var <- c(CL = 0.0182, Vc = 0.0906, Q = 0.355, Vp = 0.0874, F1 = 0.04)
rv_var  <- c(Healthy = 0.0788, Disease = 0.140)

check_var <- tibble::tibble(
  Quantity = c(names(iiv_var), names(rv_var)),
  Block = c(rep("Interindividual", length(iiv_var)), rep("Residual", length(rv_var))),
  `Table 2 variance` = c(iiv_var, rv_var),
  `Recomputed %CV` = 100 * c(sqrt(exp(iiv_var) - 1), sqrt(rv_var)),
  `Table 2 printed %CV` = c(13.6, 30.8, 65.3, 30.2, 20.2, 28.1, 37.4)
)

knitr::kable(check_var, digits = c(0, 0, 4, 1, 1),
             caption = "Both Table 2 variability columns recovered from the tabulated variances.")

stopifnot(max(abs(check_var$`Recomputed %CV` - check_var$`Table 2 printed %CV`)) < 0.06)
```

```{r params}
ui <- mod_meta
ui$iniDf[, c("name", "est", "label")] |>
  dplyr::rename("Parameter" = name, "Estimate" = est, "Label" = label) |>
  knitr::kable(digits = 4, caption = "Packaged model parameters.")
```

# Cross-model gate: reproducing the paper's own update summary

This is the sharpest available check on the transcription, and it needs no
simulation at all. Darwish 2025b Results makes two quantitative claims about how
the parameters moved when the DAFFODIL data were added to the 13-study pool:

1. "CL, Q, Vc, Vp, first-order absorption rate constant (ka), and F1 differing
   by **< 1.5%** between the two models."
2. "Covariate effect estimates were generally minimally affected (**< 10%**
   change) ...; notable exceptions were the shift in CL and Vp with positive RTT
   disease status, which **increased by 13.6% and 30.7%**, respectively,
   compared with the previous model."

Both models are packaged here, so the claims can be evaluated directly against
the two `ini()` blocks. If the update had been transcribed from the stale
footnote equations instead of from Table 2, every one of these deltas would be
exactly zero and the gate would fail.

```{r cross-model}
est_of <- function(m) setNames(m$iniDf$est, m$iniDf$name)
new <- est_of(mod_meta)
old <- est_of(prev_meta)

pct_change <- function(nm, backtransform = FALSE) {
  a <- old[[nm]]; b <- new[[nm]]
  if (backtransform) { a <- exp(a); b <- exp(b) }
  100 * (b - a) / abs(a)
}

primary <- c(CL = "lcl", Q = "lq", Vc = "lvc", Vp = "lvp", ka = "lka", F1 = "lfdepot")
cov_eff <- c("e_wt_cl", "e_crcl_cl", "e_rett_cl", "e_tbi_cl", "e_age_vc",
             "e_fxs_vc", "e_rett_vp", "e_tbi_vp", "e_fed_ka", "e_fed_f",
             "e_dose18g_f", "e_dose24g_f", "e_diarrhea_f")

primary_tbl <- tibble::tibble(
  Parameter = names(primary),
  `Darwish 2025a` = exp(old[primary]),
  `Darwish 2025b` = exp(new[primary]),
  `% change` = vapply(primary, pct_change, numeric(1), backtransform = TRUE)
)

cov_tbl <- tibble::tibble(
  `Covariate effect` = cov_eff,
  `Darwish 2025a` = old[cov_eff],
  `Darwish 2025b` = new[cov_eff],
  `% change` = vapply(cov_eff, pct_change, numeric(1))
)

knitr::kable(primary_tbl, digits = c(0, 4, 4, 2),
             caption = "Primary parameters: the paper states all six move by < 1.5%.")
knitr::kable(cov_tbl, digits = c(0, 4, 4, 2),
             caption = "Covariate effects: the paper states all move by < 10% except the two RTT shifts.")
```

```{r cross-model-gate}
# Claim 1: all six primary parameters within 1.5%.
stopifnot(nrow(primary_tbl) == 6L)
stopifnot(all(abs(primary_tbl$`% change`) < 1.5))

# Claim 2: exactly two covariate effects exceed 10%, and they are the RTT
# shifts on CL and Vp.
exceptions <- cov_tbl$`Covariate effect`[abs(cov_tbl$`% change`) >= 10]
stopifnot(identical(sort(exceptions), c("e_rett_cl", "e_rett_vp")))

# Claim 2, quantitative: those two increased by 13.6% and 30.7%.
delta <- setNames(cov_tbl$`% change`, cov_tbl$`Covariate effect`)
stopifnot(abs(delta[["e_rett_cl"]] - 13.6) < 0.1)
stopifnot(abs(delta[["e_rett_vp"]] - 30.7) < 0.1)

cat(sprintf(
  "Reproduced: RTT shift on CL %+.1f%% (paper: +13.6%%); RTT shift on Vp %+.1f%% (paper: +30.7%%).\n",
  delta[["e_rett_cl"]], delta[["e_rett_vp"]]
))
```

Both published change figures come back to three significant figures. That is
only possible if the coefficients were read from Table 2, which settles the
footnote conflict flagged at the top of this vignette.

# Virtual cohort

Individual-level data are not public (Darwish 2025b Data Availability). The
cohorts below approximate the two studies compared in the paper's Figure 3: the
DAFFODIL bands for girls aged 2-4 years, and the four LAVENDER bands for ages
5-20 years. Weight-band boundaries and doses are from Supplementary Material
Table S2.

Following the paper's own exposure simulations, all subjects have RTT, are
fasted, are free of diarrhea, and receive therapeutic (not supratherapeutic)
doses.

```{r bands}
tau <- 12 # h, BID dosing interval

bands <- tibble::tribble(
  ~study,      ~band,          ~wt_lo, ~wt_hi, ~dose_g, ~age_lo, ~age_hi, ~gfr_mean,
  "DAFFODIL",  "9-12 kg",         9,     12,      5,       2,       4,      150,
  "DAFFODIL",  "12-20 kg",       12,     20,      6,       2,       4,      150,
  "LAVENDER",  "12-20 kg (L)",   12,     20,      6,       5,       8,      162,
  "LAVENDER",  "20-35 kg",       20,     35,      8,       7,      13,      162,
  "LAVENDER",  "35-50 kg",       35,     50,     10,      11,      17,      162,
  "LAVENDER",  ">50 kg",         50,     70,     12,      14,      20,      162
) |>
  dplyr::mutate(amt_mg = dose_g * 1000)

# Keep `band` a plain character column everywhere (rxSolve `keep=`, PKNCA
# grouping, and the joins below are all type-sensitive); apply the display
# ordering only at plotting time.
band_levels <- bands$band
daffodil_bands <- bands$band[bands$study == "DAFFODIL"]
```

```{r cohort}
set.seed(20250907)
n_per_band <- 100L # well under the 200/arm cap

make_cohort <- function(band_row, n, id_offset) {
  subj <- tibble::tibble(
    id = id_offset + seq_len(n),
    study = band_row$study,
    band = band_row$band,
    amt_mg = band_row$amt_mg,
    WT = stats::runif(n, band_row$wt_lo, band_row$wt_hi),
    AGE = stats::runif(n, band_row$age_lo, band_row$age_hi),
    CRCL = pmin(pmax(stats::rnorm(n, band_row$gfr_mean, 27), 80), 220),
    DIS_RETT = 1, DIS_TBI = 0, DIS_FXS = 0,
    FED = 0, AE_DIARRHEA = 0, DOSE_18G = 0, DOSE_24G = 0
  )

  # Steady-state dose record. `ss = 1` with `ii = tau` establishes steady state
  # directly; trofinetide's disposition is slow relative to the 12 h interval,
  # so an explicit multiple-dose run would need weeks of simulated doses.
  dosing <- subj |>
    dplyr::mutate(time = 0, amt = amt_mg, evid = 1L, cmt = "depot",
                  ii = tau, ss = 1L)

  # Observations are placed on the `central` ODE state; rxode2 returns the
  # algebraic observable `Cc` as a column at those records.
  obs <- subj |>
    tidyr::crossing(time = seq(0, tau, by = 0.25)) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central",
                  ii = 0, ss = 0L)

  dplyr::bind_rows(dosing, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  lapply(seq_len(nrow(bands)), function(i) {
    make_cohort(bands[i, ], n_per_band, id_offset = (i - 1L) * n_per_band)
  })
)

stopifnot(!anyDuplicated(events[, c("id", "time", "evid")]))
stopifnot(nrow(dplyr::distinct(events, id)) == n_per_band * nrow(bands))
```

# Simulation

```{r simulate}
mod <- readModelDb("Darwish_2025b_trofinetide")

sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep = c("study", "band", "amt_mg")
) |>
  as.data.frame() |>
  # rxSolve returns observation records only (there is no `evid` column in the
  # output) and may return character keeps as factors; force them back to
  # character so the joins against the PKNCA results below are type-stable.
  dplyr::mutate(band = as.character(band), study = as.character(study))

stopifnot(!anyNA(sim$Cc), nrow(sim) > 0)
```

## Replicating Figure 4: individual steady-state profiles vs the target window

Figure 4 of Darwish 2025b overlays the 13 DAFFODIL participants' model-predicted
steady-state profiles on the average curves corresponding to AUC0-12 values of
800 and 1200 ug*h/mL, with a median curve for the subjects inside that range.
Those reference curves are, by construction, flat average concentrations of
`AUC0-12 / 12`.

```{r figure-4}
target_lines <- tibble::tibble(
  label = c("AUC0-12 = 800", "AUC0-12 = 1000", "AUC0-12 = 1200"),
  cav = c(800, 1000, 1200) / tau
)

sim |>
  dplyr::filter(band %in% daffodil_bands) |>
  dplyr::mutate(band = factor(band, levels = daffodil_bands)) |>
  dplyr::group_by(band, time) |>
  dplyr::summarise(
    Q10 = stats::quantile(Cc, 0.10, na.rm = TRUE),
    Q50 = stats::quantile(Cc, 0.50, na.rm = TRUE),
    Q90 = stats::quantile(Cc, 0.90, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  geom_hline(data = target_lines, aes(yintercept = cav, colour = label),
             linetype = "dashed") +
  facet_wrap(~band) +
  labs(
    x = "Time since dose (h)", y = "Trofinetide whole blood (ug/mL)",
    colour = NULL,
    title = "DAFFODIL steady-state profiles by weight band",
    caption = "Replicates Figure 4 of Darwish 2025b (median with 10th-90th percentile band)."
  ) +
  theme(legend.position = "bottom")
```

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, band)

# Guarantee a time = 0 record per (id, band). The ss = 1 dose means the
# time = 0 concentration is the steady-state trough, which the solve already
# produces; this bind_rows is a defensive no-op that `distinct()` collapses.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, band) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, band, time, .keep_all = TRUE) |>
  dplyr::arrange(id, band, time)

conc_obj <- PKNCA::PKNCAconc(
  sim_nca, Cc ~ time | band + id,
  concu = "ug/mL", timeu = "h"
)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, band)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | band + id, doseu = "mg")

intervals <- data.frame(
  start = 0, end = tau,
  cmax = TRUE, tmax = TRUE, ctrough = TRUE, auclast = TRUE, cav = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_auc <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(id, band, auc_pknca = PPORRES) |>
  dplyr::mutate(band = as.character(band))
stopifnot(nrow(nca_auc) == n_per_band * nrow(bands))
```

## Closed-form clearance gate

At steady state the exposure over a dosing interval reduces to
`AUC0-tau = F1 * Dose / CL`, independent of the distribution parameters. This
exercises the weight, GFR and disease-state covariate terms on CL and the whole
bioavailability chain, and it must agree with the numerically integrated PKNCA
result. Because both `CL` and `F1` carry log-normal IIV, the *median* of the
simulated AUC converges on the typical-value closed form.

```{r closed-form}
th <- setNames(ui$iniDf$est, ui$iniDf$name)

cl_typical <- function(WT, CRCL, DIS_RETT = 1, DIS_TBI = 0) {
  exp(th[["lcl"]]) * (WT / 58)^th[["e_wt_cl"]] * (CRCL / 124)^th[["e_crcl_cl"]] *
    (1 + th[["e_tbi_cl"]] * DIS_TBI) * (1 + th[["e_rett_cl"]] * DIS_RETT)
}

gate <- sim |>
  dplyr::distinct(id, study, band, WT, CRCL, amt_mg) |>
  dplyr::mutate(auc_closed_form = exp(th[["lfdepot"]]) * amt_mg / cl_typical(WT, CRCL)) |>
  dplyr::inner_join(nca_auc, by = c("id", "band")) |>
  dplyr::group_by(study, band) |>
  dplyr::summarise(
    `Closed form F1*Dose/CL` = stats::median(auc_closed_form),
    `PKNCA AUC0-12` = stats::median(auc_pknca),
    .groups = "drop"
  ) |>
  dplyr::mutate(`% diff` = 100 * (`PKNCA AUC0-12` - `Closed form F1*Dose/CL`) /
                  `Closed form F1*Dose/CL`)

knitr::kable(gate, digits = 1,
             caption = "Median closed-form vs. numerically integrated steady-state AUC0-12.")

# Both sides use the same drawn parameters, so the difference here is numerical
# (solver tolerance plus the median-of-a-ratio-of-lognormals offset), not
# physical. A tight bound is the correct assertion for this comparison.
stopifnot(nrow(gate) == nrow(bands))
stopifnot(all(abs(gate$`% diff`) < 10))
```

## Comparison against the published target exposure window

Darwish 2025b does not tabulate per-band AUC values; they appear only in the
Figure 3 box plots. What the paper *does* state numerically is its
exposure-matching criterion and its verdict: the median steady-state AUC0-12 for
each weight band must fall inside 800-1200 ug*h/mL, with 1000 ug*h/mL drawn as
the median target, and the DAFFODIL medians did. Expressing the criterion as
"within 20% of 1000 ug*h/mL" makes it exactly the tolerance the comparison table
already applies.

```{r nca_compare}
published <- tibble::tibble(
  band = daffodil_bands,
  auclast = 1000 # ug*h/mL, midpoint of the 800-1200 target window
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "band",
  params = "auclast",
  units = c(auclast = "ug*h/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated median steady-state AUC0-12 for the DAFFODIL bands vs. the 1000 ug*h/mL median target (Darwish 2025b Methods). * flags a deviation beyond the 800-1200 ug*h/mL target window."
)
```

```{r nca_compare_check}
attr(cmp, "footnote")
```

The two assertions below are evaluated on the **typical subject** of each band
rather than on the simulated cohort median. With a single log-normal eta on each
of `CL` and `F1` and an AUC that is monotone in both, the population median AUC
*is* the typical-value AUC, so nothing is lost -- and the check becomes exactly
reproducible instead of depending on which cohort the RNG happened to draw. That
matters for the second claim in particular: the band ordering is a genuine but
small (~2%) effect that a 100-subject cohort median cannot resolve, because the
Monte Carlo standard error of the median at ~24% combined CV is itself ~3%.

```{r target-gate}
band_typical <- bands |>
  dplyr::filter(study == "DAFFODIL") |>
  dplyr::mutate(
    wt_median = (wt_lo + wt_hi) / 2,
    cl = cl_typical(wt_median, gfr_mean),
    auc_typical = exp(th[["lfdepot"]]) * amt_mg / cl
  )

knitr::kable(
  band_typical[, c("band", "dose_g", "wt_median", "gfr_mean", "cl", "auc_typical")] |>
    dplyr::rename("Weight band" = band, "Dose (g)" = dose_g,
                  "Median WT (kg)" = wt_median, "GFR" = gfr_mean,
                  "Typical CL (L/h)" = cl, "Typical AUC0-12 (ug*h/mL)" = auc_typical),
  digits = c(0, 0, 1, 0, 2, 0),
  caption = "Typical-value steady-state exposure for the two DAFFODIL weight bands."
)

# Darwish 2025b Results / Conclusion: "Median steady-state AUC0-12 values for
# both body-weight bands fell within the target exposure range."
stopifnot(nrow(band_typical) == 2L)
stopifnot(all(band_typical$auc_typical >= 800), all(band_typical$auc_typical <= 1200))

# Darwish 2025b Discussion: "participants in the lower body weight group
# (>= 9 to < 12 kg) having slightly higher exposures compared with those in the
# higher body weight group". Structural under this model: the 5 g -> 6 g dose
# step (x1.20) is smaller than the clearance step implied by the 0.486 weight
# exponent over the same band midpoints ((16/10.5)^0.486 = x1.23).
lo <- band_typical$auc_typical[band_typical$band == "9-12 kg"]
hi <- band_typical$auc_typical[band_typical$band == "12-20 kg"]
stopifnot(lo > hi)

# The simulated cohort medians must land in the same place, but are asserted
# only loosely: they carry RNG-dependent Monte Carlo error.
band_median <- nca_auc |>
  dplyr::filter(band %in% daffodil_bands) |>
  dplyr::group_by(band) |>
  dplyr::summarise(median_auc = stats::median(auc_pknca), .groups = "drop") |>
  # Join by name rather than by position: PKNCA returns groups alphabetically,
  # so "12-20 kg" precedes "9-12 kg" and a positional compare would transpose.
  dplyr::inner_join(band_typical[, c("band", "auc_typical")], by = "band")
stopifnot(nrow(band_median) == 2L)
stopifnot(all(abs(band_median$median_auc - band_median$auc_typical) < 150))

cat(sprintf(
  "Typical AUC0-12,ss: 9-12 kg = %.0f, 12-20 kg = %.0f ug*h/mL (both inside 800-1200; lower band higher, ratio %.3f).\n",
  lo, hi, lo / hi
))
```

## Replicating Figure 3: distribution against the target window

```{r figure-3}
auc_by_subject <- nca_auc |>
  dplyr::left_join(dplyr::distinct(bands, study, band), by = "band") |>
  dplyr::mutate(band = factor(band, levels = band_levels))

ggplot(auc_by_subject, aes(band, auc_pknca, fill = study)) +
  geom_boxplot(outlier.alpha = 0.3) +
  geom_hline(yintercept = c(800, 1200), linetype = "dashed") +
  geom_hline(yintercept = 1000, linetype = "dotted") +
  labs(
    x = "Weight-banded dosage regimen", y = "Steady-state AUC0-12 (ug*h/mL)",
    fill = NULL,
    title = "Simulated steady-state exposure by weight band",
    caption = "Replicates Figure 3b of Darwish 2025b. Dashed lines: 800-1200 ug*h/mL target range; dotted line: 1000 ug*h/mL median target."
  ) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r figure-3-summary}
attainment <- auc_by_subject |>
  dplyr::group_by(study, band) |>
  dplyr::summarise(
    N = dplyr::n(),
    `Median AUC0-12` = stats::median(auc_pknca),
    Q1 = stats::quantile(auc_pknca, 0.25),
    Q3 = stats::quantile(auc_pknca, 0.75),
    `% within 800-1200` = 100 * mean(auc_pknca >= 800 & auc_pknca <= 1200),
    .groups = "drop"
  )

knitr::kable(attainment, digits = 1,
             caption = "Simulated steady-state AUC0-12 by weight band.")
```

The paper reports the corresponding attainment percentages as 67%
(`>=` 9 to `<` 12 kg) and 80% (`>=` 12 to `<` 20 kg) for DAFFODIL, and a range
of 57% (`>=` 50 kg) to 62% (`>=` 35 to `<` 50 kg) for LAVENDER. Those DAFFODIL
figures come from only 13 participants -- 67% and 80% are 4/6 and 4/5 -- so they
are reported here for context but are deliberately **not** asserted on: a
percentage with a denominator of five or six cannot discriminate between a
correct and a mildly incorrect transcription. The gate that does the work is the
median-within-target check above, which is the paper's own stated criterion.

# Covariate impact on steady-state exposure

Steady-state AUC over a dosing interval depends only on `F1` and `CL`, so the
model-implied exposure ratios for every covariate are available in closed form.
FXS scales the central volume alone, so it moves Cmax without touching AUC.

```{r covariate-ratios}
gmr <- tibble::tribble(
  ~Covariate,                            ~`Exposure ratio (AUC0-12)`,
  "Rett syndrome vs. healthy",           1 / (1 + th[["e_rett_cl"]]),
  "TBI vs. healthy",                     1 / (1 + th[["e_tbi_cl"]]),
  "Fragile X syndrome vs. healthy",      1,
  "Diarrhea vs. none",                   1 + th[["e_diarrhea_f"]],
  "Fed vs. fasted",                      1 + th[["e_fed_f"]],
  "18 g vs. therapeutic dose",           1 + th[["e_dose18g_f"]],
  "24 g vs. therapeutic dose",           1 + th[["e_dose24g_f"]],
  "WT 13.4 vs. 58 kg",                   (13.4 / 58)^-th[["e_wt_cl"]],
  "GFR 150 vs. 124 mL/min/1.73 m^2",     (150 / 124)^-th[["e_crcl_cl"]]
)

knitr::kable(gmr, digits = 3,
             caption = "Model-implied steady-state AUC0-12 ratios relative to the reference covariate state.")

# Structural sanity: the two supratherapeutic dose levels must reduce exposure,
# 24 g more than 18 g, and low body weight must raise it (the whole reason the
# regimen is weight-banded).
stopifnot(nrow(gmr) == 9L)
r <- setNames(gmr$`Exposure ratio (AUC0-12)`, gmr$Covariate)
stopifnot(r[["24 g vs. therapeutic dose"]] < r[["18 g vs. therapeutic dose"]])
stopifnot(r[["18 g vs. therapeutic dose"]] < 1)
stopifnot(r[["WT 13.4 vs. 58 kg"]] > 1.5)
```

The 2 kg-scale DAFFODIL participant clears trofinetide roughly half as fast as
the 58 kg reference subject, which is why a 5 g dose in a 10 kg child lands in
the same exposure window as a 12 g dose in a 50 kg adolescent.

# Assumptions and deviations

- **Coefficients come from Table 2, not from the Table 2 footnote equations.**
  The footnote's "Model equations:" block reproduces the earlier 13-study
  model's coefficients verbatim and contradicts the table it annotates. Table 2
  is used because it is the only source that reproduces the paper's own stated
  update deltas (< 1.5% on primaries; +13.6% and +30.7% on the two RTT shifts),
  verified in the cross-model gate above. Only the functional form of the
  covariate terms is taken from the footnote.
- **Age-for-weight mapping.** Darwish 2025b reports marginal covariate
  distributions per study (Table 1) but not the joint weight/age/GFR
  distribution within a weight band. Ages are drawn uniformly over a plausible
  window per band: 2-4 years for both DAFFODIL bands (the study's own
  eligibility range) and an increasing window across the LAVENDER bands. Age
  enters only the central volume, so this affects Cmax and the profile shape but
  not steady-state AUC or any assertion in this vignette.
- **GFR distribution.** Simulated as Normal(150, 27) for DAFFODIL and
  Normal(162, 27) for LAVENDER, truncated to 80-220 mL/min/1.73 m^2, matching
  the per-study means and SDs in Table 1 (ACP-2566-009: 150 (26.9);
  Neu-2566-Rett-003: 162 (36.9)). Note these paediatric GFR values sit well
  above the 124 mL/min/1.73 m^2 analysis-population median, which raises CL and
  partly offsets the weight effect.
- **Weight within a band.** Sampled uniformly across the band; the open-ended
  ">50 kg" LAVENDER band is capped at 70 kg. Real weights are not uniform, so
  the per-band medians here should be read as band-representative rather than as
  reproductions of the study's realised medians.
- **Fasted, diarrhea-free, therapeutic dose.** All simulated subjects have
  `FED = 0`, `AE_DIARRHEA = 0`, `DOSE_18G = 0`, and `DOSE_24G = 0`. DAFFODIL and
  LAVENDER dosing is not described as fed or fasted in Table 1 (both the "Fed"
  and "Fasted" counts are zero for ACP-2566-009), so the fasted reference state
  is used.
- **Steady state via `ss = 1`.** The dosing records use `ss = 1` with `ii = 12`
  rather than an explicit multi-week dosing history, which would add simulation
  time for no additional fidelity. The DAFFODIL titration through 2 g and 4 g
  BID in weeks 1-3 is therefore not simulated; the paper's exposure comparison
  is likewise made at the steady state of the final weight-banded dose.
- **Attainment percentages are not asserted.** The published 67% / 80% DAFFODIL
  figures rest on 13 participants and are too coarse to gate on; see the note
  under the Figure 3 replication.
- **Residual error is not applied to the figures.** The plots and NCA use `Cc`
  (the individual prediction) rather than a residual-error-perturbed
  observation, so they correspond to the paper's model-predicted profiles rather
  than to observed concentrations. Both published residual magnitudes are
  packaged and selected by the disease indicators; the RTT cohort simulated here
  exercises only the disease branch.
- **Bioavailability may exceed 1 for some simulated subjects.** F1 has a typical
  value of 0.832 with 20.2 %CV log-normal IIV, so the upper tail crosses 1. That
  is a property of the published parameterization, not of this implementation.
- **All parameter values come from the paper's text, tables, and supplement.**
  No value was digitized from a figure, obtained by correspondence, or carried
  from an upstream model. Weight bands, dose levels, and PK sampling schedules
  come from Supplementary Material Tables S1 and S2.
