---
title: "Darunavir (Daskapan 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Darunavir (Daskapan 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r load-model}
ui <- rxode2::rxode(readModelDb("Daskapan_2019_darunavir"))
```

# Model and source

* Citation: `r ui$reference`
* Description: `r ui$description`
* Article: <https://doi.org/10.1097/FTD.0000000000000576>
* Open-access full text: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6358182/>
* Supplement 1 (initial / prior parameter values): <http://links.lww.com/TDM/A279>
* Supplement 3 (one- and two-compartment submodel settings and AIC): <http://links.lww.com/TDM/A281>
* Supplement 4 (fat-distribution and renal-fraction grids): <http://links.lww.com/TDM/A282>

Daskapan and colleagues built a therapeutic-drug-monitoring (TDM) model for
ritonavir-boosted darunavir from routine HIV outpatient records, using the
iterative two-stage Bayesian (ITSB) procedure of the `KinPop` module of
MWPharm 3.82 rather than nonlinear mixed-effects modelling. The purpose is
narrow and explicit: translate a darunavir concentration drawn at a
convenient rather than an optimal time into a predicted trough, so the trough
can be scored against the local follow-up cut-offs.

Two things follow from that purpose and shape everything below.

First, the reported structure is deliberately small. A one-compartment model
with first-order absorption and elimination won; adding a second compartment
did not improve the AIC and estimated a peripheral volume of 0.051 L/kg,
"which is negligible as a significant peripheral compartment" (Results).
Absorption could not be estimated at all, because few samples fell in the
0-4 h window, so `Ka` was held at a literature value.

Second, the published validation is an agreement analysis between measured and
individually-Bayesian-predicted concentrations (Passing-Bablok regression in
Figure 1, Bland-Altman in Figure 2). Those figures are functions of the
individual patient records, which are not public, so **they cannot be
reproduced here** and no attempt is made to. What this vignette does instead is
verify that the packaged model *is* the published model: every derived quantity
is checked against the paper's own equations, the solved system is checked
against its own closed form, and each fixed structural switch is checked with a
mutation control that proves the term is live rather than dead code.

# Population

The development set is "hospital A" -- ASST Fatebenefratelli Sacco University
Hospital, Milano -- a retrospective record review over April 2015 to August
2017. `r ui$population$n_subjects` unique adult patients contributed
`r ui$population$n_observations` darunavir plasma samples, i.e. **one sample per
patient**. Median age was 54 years (range 24-74), median weight 72.0 kg (range
40-123), median height 173.0 cm (range 150-193), median BMI 24.6 kg/m^2 (range
16.9-35.3) and median serum creatinine 83.5 umol/L (range 44.2-230.7); 29% were
women (Table 1). All patients received darunavir/ritonavir 800/100 mg once
daily (162 of 198, 82%) or 600/100 mg twice daily (36 of 198, 18%). Samples
were excluded when the intake or sampling time was unknown, or when the
concentration fell below the 0.2 mg/L lower limit of quantification. Race and
ethnicity are not reported.

A separate 170-patient / 170-sample set from University Medical Center
Groningen ("hospital B", January 2010 to May 2017) served only as the external
validator and contributed nothing to parameter estimation.

The same information is available programmatically:

```{r population}
str(ui$population)
```

# Source trace

Per-parameter provenance is recorded as an in-file comment beside each `ini()`
entry in `inst/modeldb/specificDrugs/Daskapan_2019_darunavir.R`. Collected here
for review. "Model 2" is the column of Table 2 that footnote `*` marks as the
chosen final model.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` (`Ka`) | 1.04 1/h, fixed | Table 2, Model 2, `Ka`; footnotes "literature value" (Arab-Alameddine 2014) and "set on fixed value". Supplement 1 one-compartment prior `Ka` 1.04 (0.35) |
| `lcl_nonren` (`CLm`) | 9.47 L/h/70 kg | Table 2, Model 2, `CLm` mean (95% CI 8.24-10.65) |
| `lvd` (`Vd`) | 2.13 L/kg LBMc | Table 2, Model 2, `Vd` mean (95% CI 1.39-3.26) |
| `lfdepot` (`F`) | 0.82, fixed | Table 2, `F`; footnote "literature value from SPC". Results: "bioavailability was fixed in all parameterizations at the literature value of 0.82" |
| `e_wt_vd` (`fd`) | 5, fixed | Table 2, "Fat distribution". Supplement 4 AIC grid 1:1 = 3743.29, 1:2 = 3676.84, 1:5 = 3603.29, 1:10 = 3630.27, 1:20 = 3602.20 |
| `e_crcl_cl_renal` (`fr`) | 0, fixed | Table 2, Model 2, `fr`. Supplement 4: `fr` = 0.12 gives AIC 1611.08, `fr` = 0 gives 1584.89 |
| `etalcl_nonren` | 0.355749 | Table 2, Model 2, `CLm` SD 6.19 (4.85-7.76) -> CV 65.4% -> `log(1 + CV^2)` |
| `etalvd` | 0.912284 | Table 2, Model 2, `Vd` SD 2.60 (1.43-4.66) -> CV 122.1% -> `log(1 + CV^2)` |
| `addSd`, `propSd` | 0.2 mg/L, 0.05, fixed | Methods: "the residual error was assumed to be normally distributed and equal to the SD of the assay, which was estimated as 0.2 + 0.05 * C" |
| `cl = CLm * (WT / 70) + fr * CLcr` | n/a | Methods, Population Pharmacokinetic Model Development, CL equation |
| `vc = Vd * LBMc` | n/a | Methods, same paragraph, V equation |
| `LBMc = LBM + (WT - LBM) * fd` | n/a | Methods, same paragraph |
| `LBM = 50.0 + 0.9 * (HT - 152)` (men), `45.5 + 0.9 * (HT - 152)` (women) | n/a | Methods, same paragraph |
| 1-compartment ODEs, first-order absorption and elimination | n/a | Results: "A 1-compartment model with a first-order absorption and elimination, a distribution to fatty tissue factor (fd) of 5, and a fr value of zero resulted in the best model" |

```{r ini-table}
ui$iniDf |>
  dplyr::select(name, est, fix, label) |>
  dplyr::rename(
    "Parameter" = name, "Estimate" = est, "Fixed" = fix, "Label" = label
  ) |>
  knitr::kable(
    digits = 6,
    caption = "Packaged ini() block. `Fixed = TRUE` marks values the source held constant rather than estimating."
  )
```

# Structural verification

These checks are deterministic: each one re-derives a published quantity
independently and compares it with what the compiled model actually computes.
None of them depends on a random draw, so the tolerances are numerical rather
than statistical.

## The solved system keeps both compartments

rxode2 can match a `cl` / `vc` pair against its analytic kernel and discard the
explicit `d/dt` bodies. For a one-compartment oral model the analytic and ODE
systems are identical, so this would not change the predictions, but it would
drop the state columns. Assert that it did not happen.

```{r gate-states}
typ <- rxode2::zeroRe(ui)

cov_grid <- tibble::tibble(
  id   = 1:12,
  SEXF = c(0, 0, 0, 0, 0, 0, 1, 1, 1, 1, 1, 1),
  HT   = c(160, 173, 173, 180, 190, 193, 152, 165, 165, 170, 178, 185),
  WT   = c( 62,  60,  72,  95, 110, 123,  48,  55,  65,  80,  95, 105),
  CRCL = c( 60,  90, 100, 110,  90,  75,  55,  90, 100, 120,  85,  70)
)

sd_grid <- dplyr::bind_rows(
  cov_grid |> dplyr::mutate(time = 0, amt = 800, evid = 1L, cmt = "depot"),
  cov_grid |>
    tidyr::crossing(time = sort(unique(c(
      0, exp(seq(log(0.01), log(24), length.out = 220)),
      seq(24, 300, by = 0.5)
    )))) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
) |>
  dplyr::arrange(id, time, evid)

sd_typ <- rxode2::rxSolve(
  typ, sd_grid,
  omega = NA, # mandatory: rxSolve reuses a previous solve's omega otherwise
  keep = c("WT", "HT", "SEXF", "CRCL"),
  returnType = "data.frame"
)

stopifnot(
  all(c("depot", "central") %in% names(sd_typ)),
  # zeroRe actually took: one cl per covariate row, not a random draw
  nrow(dplyr::distinct(sd_typ, id, cl_nonren)) == nrow(cov_grid)
)
```

## Every derived covariate term matches the published equations

`LBM`, `LBMc`, the clearance arms and the absolute volume are recomputed here in
plain R straight from the Methods equations, then compared with the columns the
compiled model returns.

```{r gate-covariates}
derived <- sd_typ |>
  dplyr::distinct(id, WT, HT, SEXF, CRCL, lbm, lbmc, cl_nonren, cl_renal, cl, vd, vc) |>
  dplyr::mutate(
    # Methods, V equation
    lbm_ref  = ifelse(SEXF == 1, 45.5, 50.0) + 0.9 * (HT - 152),
    lbmc_ref = lbm_ref + (WT - lbm_ref) * 5,
    vc_ref   = 2.13 * lbmc_ref,
    # Methods, CL equation, with fr = 0 and CLcr converted mL/min -> L/h
    cl_nonren_ref = 9.47 * (WT / 70),
    cl_renal_ref  = 0 * 0.06 * CRCL,
    cl_ref        = cl_nonren_ref + cl_renal_ref
  )

max_rel <- function(a, b) max(abs(a - b) / pmax(abs(b), 1e-12))

stopifnot(
  max_rel(derived$lbm, derived$lbm_ref) < 1e-12,
  max_rel(derived$lbmc, derived$lbmc_ref) < 1e-12,
  max_rel(derived$vc, derived$vc_ref) < 1e-10,
  max_rel(derived$cl_nonren, derived$cl_nonren_ref) < 1e-10,
  max_rel(derived$cl, derived$cl_ref) < 1e-10,
  all(derived$cl_renal == 0)
)

derived |>
  dplyr::transmute(
    SEXF, HT, WT,
    LBM = round(lbm, 2), LBMc = round(lbmc, 2),
    `CL (L/h)` = round(cl, 3), `Vc (L)` = round(vc, 1),
    `t1/2 (h)` = round(log(2) * vc / cl, 1)
  ) |>
  dplyr::rename("Female" = SEXF, "Height (cm)" = HT, "Weight (kg)" = WT) |>
  knitr::kable(caption = "Derived individual parameters over a representative covariate grid.")
```

## The solved profile matches its own closed form

A one-compartment model with first-order absorption has the Bateman solution

`Cc(t) = F * Dose * ka / (vc * (ka - kel)) * (exp(-kel * t) - exp(-ka * t))`.

Both sides use each subject's own solved `ka`, `cl` and `vc`, so the only
difference admissible here is solver quadrature error. This is the check that
would go red if the compiled system were structurally different from the
published one.

```{r gate-closed-form}
cf <- sd_typ |>
  dplyr::filter(!is.na(Cc), time > 0) |>
  dplyr::mutate(
    kel = cl / vc,
    Cc_closed = exp(ui$theta[["lfdepot"]]) * 800 * ka / (vc * (ka - kel)) *
      (exp(-kel * time) - exp(-ka * time))
  ) |>
  # Floor the comparison at 1e-6 mg/L. The grid runs to 300 h, and in the far
  # tail both sides are 1e-19 mg/L -- nineteen orders of magnitude below the
  # paper's 0.2 mg/L limit of quantification -- where the solver's ABSOLUTE
  # tolerance dominates and a RELATIVE difference is meaningless (it reaches
  # 0.70 there). 1e-6 mg/L is still five orders below anything measurable.
  dplyr::filter(Cc_closed > 1e-6)

stopifnot(
  nrow(cf) > 2000, # the gate has rows to test
  # Deterministic quantity (no cohort, no RNG), so the bound is tight: the
  # realised maximum is 7.1e-7, set by solver quadrature, not by a draw.
  max_rel(cf$Cc, cf$Cc_closed) < 1e-5
)
cat(sprintf(
  "closed-form agreement: max relative difference %.2e over %d points\n",
  max_rel(cf$Cc, cf$Cc_closed), nrow(cf)
))
```

## Mutation controls -- each fixed switch is live, not dead code

Three of the model's parameters are fixed at values that make their term
disappear or collapse (`fr` = 0 removes the renal clearance arm entirely,
`fd` = 5 is a grid-selected constant, `F` = 0.82 is a literature anchor). A
gate that only checks the published configuration cannot distinguish a wired
term from a mis-typed one. Each is therefore perturbed and the predicted
consequence asserted exactly.

```{r gate-mutations}
one <- sd_grid |> dplyr::filter(id == 3) # 72 kg, 173 cm man, CRCL 100

solve_one <- function(params = NULL) {
  rxode2::rxSolve(
    typ, one, omega = NA, params = params,
    keep = c("WT", "HT", "SEXF", "CRCL"), returnType = "data.frame"
  )
}

base <- solve_one()
base_cl <- base$cl[1]
base_vc <- base$vc[1]

# (1) fr: restoring the SPC renal fraction of 0.12 that Supplement 4 rejected
#     must add exactly fr * CLcr(L/h) to total clearance.
mut_fr <- solve_one(c(e_crcl_cl_renal = 0.12))
expect_fr <- base_cl + 0.12 * 0.06 * 100

# (2) fd: fd = 1 collapses LBMc to total body weight, by construction
#     (LBMc = LBM + (WT - LBM) * 1 = WT). An algebraic identity, so exact.
mut_fd <- solve_one(c(e_wt_vd = 1))

# (3) F: setting bioavailability to 1 must scale every concentration by 1/0.82.
mut_f <- solve_one(c(lfdepot = log(1)))
both <- dplyr::inner_join(
  # Same 1e-6 mg/L floor as the closed-form gate, for the same reason.
  base |> dplyr::filter(!is.na(Cc), Cc > 1e-6) |> dplyr::select(time, Cc),
  mut_f |> dplyr::filter(!is.na(Cc), Cc > 1e-6) |> dplyr::select(time, Cc_f = Cc),
  by = "time"
)

stopifnot(
  # each mutation moved something
  abs(mut_fr$cl[1] - base_cl) > 1e-6,
  abs(mut_fd$vc[1] - base_vc) > 1e-6,
  nrow(both) > 500, # the single-subject grid holds ~770 observation times
  # and moved it by exactly the predicted amount
  abs(mut_fr$cl[1] - expect_fr) < 1e-9,
  abs(mut_fd$lbmc[1] - 72) < 1e-9,
  max_rel(both$Cc_f / both$Cc, rep(1 / 0.82, nrow(both))) < 1e-5
)

tibble::tibble(
  Mutation = c("fr: 0 -> 0.12", "fd: 5 -> 1", "F: 0.82 -> 1"),
  Quantity = c("CL (L/h)", "LBMc (kg)", "Cc ratio vs published"),
  Published = c(round(base_cl, 4), round(base$lbmc[1], 2), 1),
  Mutated = c(round(mut_fr$cl[1], 4), round(mut_fd$lbmc[1], 2), round(mean(both$Cc_f / both$Cc), 4)),
  Predicted = c(round(expect_fr, 4), 72, round(1 / 0.82, 4))
) |>
  knitr::kable(caption = "Mutation controls. Each fixed switch changes the model by exactly the amount its equation predicts.")
```

## Dose recovery through PKNCA

For a linear one-compartment model the clearance identity `CL * AUC(0-inf) =
F * Dose` holds exactly. AUC is computed with PKNCA over the covariate grid,
never by an inline trapezoid. The same identity re-run with `F` mutated to 1 is
the control that keeps the gate from passing vacuously.

```{r gate-auc-recovery}
nca_recovery <- function(sim, dose = 800) {
  conc <- sim |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::transmute(id, time, Cc = pmax(Cc, 0), regimen = "single dose")
  conc <- dplyr::bind_rows(
    conc,
    conc |> dplyr::distinct(id, regimen) |> dplyr::mutate(time = 0, Cc = 0)
  ) |>
    dplyr::distinct(id, regimen, time, .keep_all = TRUE) |>
    dplyr::arrange(id, time)

  dose_df <- sim |>
    dplyr::distinct(id) |>
    dplyr::mutate(time = 0, amt = dose, regimen = "single dose")

  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(as.data.frame(conc), Cc ~ time | regimen + id,
      concu = "mg/L", timeu = "h"
    ),
    PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | regimen + id,
      doseu = "mg", route = "extravascular"
    ),
    intervals = data.frame(
      start = 0, end = Inf,
      cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
    )
  ))
  as.data.frame(res$result) |>
    dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
    dplyr::select(id, PPTESTCD, PPORRES) |>
    tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
}

rec_base <- nca_recovery(sd_typ) |>
  dplyr::left_join(dplyr::select(derived, id, cl, vc), by = "id") |>
  dplyr::mutate(recovered = cl * aucinf.obs / 800)

sd_f1 <- rxode2::rxSolve(
  typ, sd_grid,
  omega = NA, params = c(lfdepot = log(1)),
  keep = c("WT", "HT", "SEXF", "CRCL"), returnType = "data.frame"
)
rec_f1 <- nca_recovery(sd_f1) |>
  dplyr::left_join(dplyr::select(derived, id, cl), by = "id") |>
  dplyr::mutate(recovered = cl * aucinf.obs / 800)

stopifnot(
  nrow(rec_base) == nrow(cov_grid), # the gate has rows to test
  !anyNA(rec_base$aucinf.obs),
  # published F = 0.82
  max(abs(rec_base$recovered - 0.82)) < 0.005,
  # mutation control: with F = 1 the same identity must recover 1, not 0.82
  max(abs(rec_f1$recovered - 1.00)) < 0.005,
  # and PKNCA's half-life must match log(2) * vc / cl
  max_rel(rec_base$half.life, log(2) * rec_base$vc / rec_base$cl) < 0.01
)

rec_base |>
  dplyr::transmute(
    id,
    `Cmax (mg/L)` = round(cmax, 3),
    `Tmax (h)` = round(tmax, 2),
    `AUC0-inf (mg*h/L)` = round(aucinf.obs, 1),
    `t1/2 (h)` = round(half.life, 1),
    `CL * AUC / Dose` = round(recovered, 4)
  ) |>
  knitr::kable(caption = "Single 800 mg dose, typical values over the covariate grid. The last column recovers the fixed bioavailability of 0.82.")
```

# The `fd = 5` domain bound

This is the one property of the published parameterisation that a user must
know before simulating it, and it is a consequence of the fitted value rather
than of the transcription.

With `fd = 1` the corrected lean body mass `LBMc = LBM + (WT - LBM) * fd`
equals total body weight and with `fd = 0` it equals lean body mass; the
selected `fd = 5` therefore **extrapolates** rather than interpolating, and
rearranges to `LBMc = 5 * WT - 4 * LBM`. That is positive only when
`WT > 0.8 * LBM`, i.e.

* men: `WT > 40 + 0.72 * (HT - 152)` kg,
* women: `WT > 36.4 + 0.72 * (HT - 152)` kg.

A 173 cm man needs more than 55.1 kg and a 193 cm man more than 69.5 kg. Below
the bound `LBMc`, and hence the volume of distribution, is negative. The bound
is reachable inside the source cohort's own reported ranges, which extend to
193 cm, down to 40 kg and down to a BMI of 16.9 kg/m^2: a 173 cm man at
BMI 16.9 weighs 50.6 kg and falls below it. Any virtual cohort built from this
model must therefore be restricted, which is what the next section does.

```{r domain-bound, fig.height = 4}
bound <- tidyr::crossing(
  HT = seq(150, 195, by = 1),
  SEXF = c(0, 1)
) |>
  dplyr::mutate(
    LBM = ifelse(SEXF == 1, 45.5, 50.0) + 0.9 * (HT - 152),
    WT_min = 0.8 * LBM,
    Sex = ifelse(SEXF == 1, "Women", "Men")
  )

# Verified against the model itself rather than against the algebra alone.
stopifnot(
  abs(with(dplyr::filter(bound, Sex == "Men", HT == 173), WT_min) - 55.12) < 0.01,
  abs(with(dplyr::filter(bound, Sex == "Men", HT == 193), WT_min) - 69.52) < 0.01
)

ggplot(bound, aes(HT, WT_min, colour = Sex)) +
  geom_line(linewidth = 0.9) +
  geom_ribbon(aes(ymin = -Inf, ymax = WT_min, fill = Sex), alpha = 0.12, colour = NA) +
  annotate("point", x = 173, y = 50.6, shape = 4, size = 3) +
  annotate("text",
    x = 173, y = 47, hjust = 0.5, size = 3,
    label = "173 cm man at the cohort's\nminimum BMI of 16.9 kg/m^2"
  ) +
  labs(
    x = "Height (cm)", y = "Minimum body weight for a positive LBMc (kg)",
    title = "Domain bound created by the fitted fat-distribution factor fd = 5",
    caption = "Shaded region: LBMc = 5 * WT - 4 * LBM is negative, so the model returns a negative volume."
  ) +
  theme_bw()
```

# Virtual cohort

Individual patient data are not public. The cohort below approximates the
Table 1 demographics of hospital A -- 29% women, median BMI 24.6 kg/m^2 (range
16.9-35.3), sex-appropriate heights -- and is then restricted to the
positive-`LBMc` domain above. The number of draws that restriction removes is
reported rather than silently absorbed.

```{r cohort}
# set.seed() seeds R's RNG, which is what draws the covariates below, so the
# COHORT is reproducible. It does NOT seed rxode2's eta sampler, whose streams
# are partitioned per solver thread -- so every assertion on an eta-driven
# quantity downstream is written to hold for any draw. See pattern 12 of
# references/known-vignette-failure-patterns.md.
set.seed(20190201)

n_per_arm <- 200L

draw_subjects <- function(n_draw) {
  sexf <- rbinom(n_draw, 1L, 0.29)
  ht <- round(ifelse(sexf == 1L, rnorm(n_draw, 163, 6.5), rnorm(n_draw, 176, 7)))
  ht <- pmin(pmax(ht, 150), 193) # Table 1 height range
  bmi <- pmin(pmax(rnorm(n_draw, 24.6, 3.6), 16.9), 35.3) # Table 1 BMI range
  tibble::tibble(
    SEXF = sexf, HT = ht,
    WT = round(pmin(pmax(bmi * (ht / 100)^2, 40), 123), 1), # Table 1 weight range
    # CKD-EPI eGFR; enters only through fr, which the final model fixes at 0.
    CRCL = round(pmin(pmax(rnorm(n_draw, 95, 22), 25), 150), 1)
  ) |>
    dplyr::mutate(
      LBM = ifelse(SEXF == 1L, 45.5, 50.0) + 0.9 * (HT - 152),
      in_domain = WT > 0.8 * LBM
    )
}

pool <- draw_subjects(3000L)
n_dropped <- sum(!pool$in_domain)
cat(sprintf(
  "domain restriction: %d of %d draws (%.1f%%) had WT <= 0.8 * LBM and were dropped\n",
  n_dropped, nrow(pool), 100 * n_dropped / nrow(pool)
))
stopifnot(n_dropped > 0) # the restriction is doing work, not decorative

subjects <- pool |>
  dplyr::filter(in_domain) |>
  head(2L * n_per_arm) |>
  dplyr::mutate(id = seq_len(dplyr::n())) |>
  dplyr::select(id, SEXF, HT, WT, CRCL, LBM)

stopifnot(nrow(subjects) == 2L * n_per_arm)

# 40 days of dosing, not the 14 that a median half-life of ~13 h would suggest.
# The 122% CV on Vd puts individual half-lives as high as ~480 h, and a subject
# who has not accumulated to steady state has an AUC over the final interval
# well below F * Dose / CL -- measured at 14 days, 10% of a 100-subject draw
# failed the identity, down to 38% recovery. Dose events are cheap; only
# observation rows cost, and those are confined to the first and last intervals.
regimens <- tibble::tribble(
  ~regimen,             ~amt, ~ii, ~n_dose,
  "600/100 mg q12h",     600,  12,     80L,
  "800/100 mg q24h",     800,  24,     40L
)

t_end <- 960 # h = 40 days; identical for both arms

make_arm <- function(reg, subj) {
  last_dose <- reg$ii * (reg$n_dose - 1L)
  obs_times <- sort(unique(c(
    seq(0, 24, by = 0.5), # first dosing interval, for the accumulation view
    last_dose, # exact interval start
    # Log-spaced immediately after the dose so the sharp early peak of a
    # low-volume subject is resolved, then uniform across the interval.
    last_dose + exp(seq(log(0.01), log(reg$ii), length.out = 120)),
    seq(last_dose, t_end, length.out = 97) # includes BOTH endpoints exactly,
  ))) # which cmin needs
  dplyr::bind_rows(
    subj |>
      tidyr::crossing(time = reg$ii * seq(0L, reg$n_dose - 1L)) |>
      dplyr::mutate(amt = reg$amt, evid = 1L, cmt = "depot"),
    subj |>
      tidyr::crossing(time = obs_times) |>
      dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
  ) |>
    dplyr::mutate(regimen = reg$regimen, ii_h = reg$ii, last_dose = last_dose)
}

events <- dplyr::bind_rows(
  make_arm(regimens[1, ], subjects[1:n_per_arm, ]),
  make_arm(regimens[2, ], subjects[(n_per_arm + 1):(2 * n_per_arm), ])
) |>
  dplyr::arrange(id, time, evid)

# IDs are disjoint across arms, so no subject can silently receive both.
stopifnot(
  !anyDuplicated(unique(events[, c("id", "time", "evid")])),
  length(intersect(
    unique(events$id[events$regimen == regimens$regimen[1]]),
    unique(events$id[events$regimen == regimens$regimen[2]])
  )) == 0L,
  all(subjects$WT > 0.8 * subjects$LBM),
  all(subjects$LBM + (subjects$WT - subjects$LBM) * 5 > 0)
)
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  ui, events,
  omega = ui$omega, # mandatory: see the note in the cohort chunk
  keep = c("regimen", "WT", "HT", "SEXF", "CRCL", "ii_h", "last_dose"),
  returnType = "data.frame"
)

stopifnot(
  all(c("depot", "central") %in% names(sim)),
  # the population solve really did sample IIV (the reverse of the omega trap)
  dplyr::n_distinct(round(sim$vd, 8)) > 100L,
  dplyr::n_distinct(round(sim$cl_nonren, 8)) > 100L,
  # and the drawn volumes stayed positive over the restricted cohort
  all(sim$vc > 0), all(sim$cl > 0), !anyNA(sim$Cc)
)
```

The cohort's realised variability is a draw, so the check below is deliberately
loose: it confirms the `omega^2 = log(1 + CV^2)` encoding put the right order of
magnitude of spread on each parameter, not that any particular draw reproduced
the published SD.

```{r iiv-recovery}
iiv <- sim |>
  dplyr::distinct(id, cl_nonren, vd) |>
  dplyr::summarise(
    `CLm median (L/h/70 kg)` = median(cl_nonren),
    `CLm CV (%)` = 100 * sd(cl_nonren) / mean(cl_nonren),
    `Vd median (L/kg LBMc)` = median(vd),
    `Vd CV (%)` = 100 * sd(vd) / mean(vd)
  )

# Published: CLm 9.47 with SD 6.19 (CV 65.4 %); Vd 2.13 with SD 2.60 (CV 122.1 %).
# The MEDIAN of a log-normal whose median is the published mean is the published
# mean, so those are checked with generous multiplicative bounds; the CVs are
# checked only for order of magnitude because a 400-subject draw of a CV-122 %
# log-normal has a very wide sampling distribution.
stopifnot(
  iiv$`CLm median (L/h/70 kg)` > 9.47 / 1.5, iiv$`CLm median (L/h/70 kg)` < 9.47 * 1.5,
  iiv$`Vd median (L/kg LBMc)` > 2.13 / 1.5, iiv$`Vd median (L/kg LBMc)` < 2.13 * 1.5,
  iiv$`CLm CV (%)` > 35, iiv$`CLm CV (%)` < 110,
  iiv$`Vd CV (%)` > 60, iiv$`Vd CV (%)` < 220
)

iiv |>
  dplyr::mutate(dplyr::across(dplyr::everything(), \(x) round(x, 2))) |>
  knitr::kable(caption = "Realised cohort variability against the published CLm 9.47 (SD 6.19) and Vd 2.13 (SD 2.60).")
```

# Simulated concentration-time behaviour

The paper publishes no concentration-versus-time figure -- its Figures 1 and 2
are Passing-Bablok and Bland-Altman agreement plots built from individual
patient records, which are not public. The panels below therefore characterise
the packaged model rather than replicating a published figure.

```{r vpc-first-interval}
sim |>
  dplyr::filter(!is.na(Cc), time <= 24) |>
  dplyr::group_by(regimen, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05), Q50 = quantile(Cc, 0.50), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(linewidth = 0.8) +
  facet_wrap(~regimen) +
  scale_x_continuous(breaks = seq(0, 24, 6)) +
  labs(
    x = "Time after first dose (h)", y = "Darunavir concentration (mg/L)",
    title = "First 24 h after treatment start",
    caption = "Median with 5th-95th percentile band, 200 virtual patients per regimen."
  ) +
  theme_bw()
```

```{r vpc-steady-state}
sim |>
  dplyr::filter(!is.na(Cc), time >= last_dose) |>
  dplyr::mutate(tad = time - last_dose) |>
  dplyr::group_by(regimen, tad) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05), Q50 = quantile(Cc, 0.50), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "darkorange") +
  geom_line(linewidth = 0.8) +
  geom_hline(
    data = tibble::tibble(
      regimen = regimens$regimen, cutoff = c(2.60, 1.07)
    ),
    aes(yintercept = cutoff), linetype = "dashed"
  ) +
  facet_wrap(~regimen, scales = "free_x") +
  labs(
    x = "Time after the last dose (h)", y = "Darunavir concentration (mg/L)",
    title = "Steady state, final dosing interval",
    caption = paste(
      "Median with 5th-95th percentile band. Dashed line: the follow-up trough",
      "cut-off quoted in the Discussion (1.07 mg/L once daily, 2.60 mg/L twice daily)."
    )
  ) +
  theme_bw()
```

# PKNCA validation

Steady-state NCA over the final dosing interval of each regimen, per PKNCA
recipe 3. The observation grid places records at both interval endpoints
exactly, which `cmin` needs.

Note the `regimen` column on the `intervals` frame. Without it PKNCA applies
**every** interval row to **every** group, so each subject would receive results
for both the q12h and the q24h window; `pivot_wider()` then silently produces
list columns and an `anyNA()` check on them returns `FALSE` regardless of
content -- a gate that cannot go red. The type assertions below exist for that
reason.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::transmute(id, time, Cc, regimen = as.character(regimen))

# Guarantee a time = 0 record per subject; for extravascular dosing the
# pre-dose concentration is 0.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, regimen) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, regimen, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

dose_df <- events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, regimen) |>
  dplyr::mutate(regimen = as.character(regimen))

conc_obj <- PKNCA::PKNCAconc(
  as.data.frame(sim_nca), Cc ~ time | regimen + id,
  concu = "mg/L", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  as.data.frame(dose_df), amt ~ time | regimen + id,
  doseu = "mg", route = "extravascular"
)

ss_windows <- events |>
  dplyr::distinct(regimen, ii_h, last_dose) |>
  dplyr::arrange(regimen) # PKNCA orders groups alphabetically

intervals <- data.frame(
  regimen = as.character(ss_windows$regimen), # binds each window to its own arm
  start = ss_windows$last_dose,
  end = ss_windows$last_dose + ss_windows$ii_h,
  cmax = TRUE, tmax = TRUE, cmin = TRUE,
  auclast = TRUE, cav = TRUE,
  stringsAsFactors = FALSE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_par <- c("cmax", "tmax", "cmin", "auclast", "cav")
nca_raw <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% nca_par)

# One result per subject per parameter -- proves the interval grouping worked
# and that the pivot below cannot produce a list column.
stopifnot(
  nrow(dplyr::count(nca_raw, regimen, id, PPTESTCD) |> dplyr::filter(n > 1L)) == 0L
)

nca_tbl <- nca_raw |>
  dplyr::select(regimen, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

stopifnot(
  nrow(nca_tbl) == 2L * n_per_arm, # every subject produced a result
  # Type guard: a list column would make every anyNA() below pass vacuously.
  all(vapply(nca_tbl[nca_par], is.numeric, logical(1))),
  !anyNA(nca_tbl$auclast), !anyNA(nca_tbl$cmin), !anyNA(nca_tbl$cmax)
)
```

At steady state a linear model must satisfy `CL * AUC(0-tau) = F * Dose`,
evaluated per subject against that subject's own drawn clearance. It is exact up
to quadrature, so it is a structural gate rather than a statistical one, and it
is the check that would catch a lost dose event, a wrong `f(depot)` target or a
mis-scaled interval.

The identity only holds **once a subject has actually accumulated to steady
state**, which is not automatic here: the 122% CV on `Vd` gives some subjects
half-lives of several hundred hours. Each subject's attainment is therefore
tested explicitly against its own half-life and the shortfall is reported rather
than absorbed into a loosened tolerance.

```{r gate-ss-mass-balance}
subject_pk <- sim |>
  dplyr::distinct(id, regimen, cl, vc) |>
  dplyr::mutate(
    regimen = as.character(regimen),
    thalf = log(2) * vc / cl,
    # 7 half-lives leaves under 1% of the approach to steady state outstanding.
    ss_attained = t_end >= 7 * thalf
  )
stopifnot(nrow(subject_pk) == 2L * n_per_arm) # one row per subject, no fan-out

mb <- nca_tbl |>
  dplyr::left_join(subject_pk, by = c("id", "regimen")) |>
  dplyr::left_join(regimens |> dplyr::select(regimen, dose_amt = amt), by = "regimen") |>
  dplyr::mutate(recovered = cl * auclast / (0.82 * dose_amt))

stopifnot(nrow(mb) == 2L * n_per_arm, !anyNA(mb$recovered), !anyNA(mb$ss_attained))

n_not_ss <- sum(!mb$ss_attained)
cat(sprintf(
  paste0(
    "half-life: median %.1f h, max %.1f h\n",
    "not yet at steady state after %d h (t1/2 > %.0f h): %d of %d subjects\n",
    "CL * AUCtau / (F * Dose), steady-state subjects only: %.4f - %.4f\n"
  ),
  median(mb$thalf), max(mb$thalf), t_end, t_end / 7, n_not_ss, nrow(mb),
  min(mb$recovered[mb$ss_attained]), max(mb$recovered[mb$ss_attained])
))

stopifnot(
  # The gate must have most of the cohort to test.
  sum(mb$ss_attained) > 0.9 * nrow(mb),
  # Quadrature over one interval plus the < 1% of accumulation still outstanding
  # at 7 half-lives. Observed maxima across draws sat at or under 1.0%; 3%
  # leaves headroom for which subjects land near the attainment boundary while
  # still going red on a structural break (dropping f(depot) moves this 22%).
  max(abs(mb$recovered[mb$ss_attained] - 1)) < 0.03
)
```

The same identity on the deterministic typical-value covariate grid, where every
subject is comfortably at steady state, is where the tight bound belongs.

```{r gate-ss-typical}
ss_typ <- lapply(seq_len(nrow(regimens)), function(i) {
  reg <- regimens[i, ]
  ld <- reg$ii * (reg$n_dose - 1L)
  ev <- dplyr::bind_rows(
    cov_grid |> tidyr::crossing(time = reg$ii * seq(0L, reg$n_dose - 1L)) |>
      dplyr::mutate(amt = reg$amt, evid = 1L, cmt = "depot"),
    cov_grid |> tidyr::crossing(time = sort(unique(c(
      ld, ld + exp(seq(log(0.01), log(reg$ii), length.out = 150)),
      seq(ld, ld + reg$ii, length.out = 97)
    )))) |>
      dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
  ) |>
    dplyr::arrange(id, time, evid)
  s <- rxode2::rxSolve(typ, ev, omega = NA, returnType = "data.frame")
  s <- s[!is.na(s$Cc) & s$time >= ld, ]
  # pk.calc.auc.last() defaults its interval to start at 0, so it returns NA on
  # an absolute-time window starting at 936 h. Pass time after the dose instead.
  s$tad <- s$time - ld
  s |>
    dplyr::group_by(id) |>
    dplyr::summarise(
      cl = dplyr::first(cl),
      auctau = PKNCA::pk.calc.auc.last(conc = Cc, time = tad),
      .groups = "drop"
    ) |>
    dplyr::mutate(regimen = reg$regimen, recovered = cl * auctau / (0.82 * reg$amt))
}) |>
  dplyr::bind_rows()

stopifnot(
  nrow(ss_typ) == 2L * nrow(cov_grid),
  !anyNA(ss_typ$recovered),
  # Deterministic: quadrature only. Realised maximum deviation 0.03%.
  max(abs(ss_typ$recovered - 1)) < 0.005
)
cat(sprintf(
  "typical-value steady-state recovery: %.5f - %.5f over %d covariate/regimen combinations\n",
  min(ss_typ$recovered), max(ss_typ$recovered), nrow(ss_typ)
))
```

```{r nca-summary}
nca_summary <- nca_tbl |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(
    n = dplyr::n(),
    cmax = median(cmax), tmax = median(tmax),
    cmin = median(cmin), cav = median(cav), auclast = median(auclast),
    cmin_p05 = quantile(cmin, 0.05), cmin_p95 = quantile(cmin, 0.95),
    .groups = "drop"
  )

nca_summary |>
  dplyr::transmute(
    regimen, n,
    cmax = round(cmax, 2), tmax = round(tmax, 2),
    cmin = round(cmin, 2),
    cmin_range = sprintf("%.2f - %.2f", cmin_p05, cmin_p95),
    cav = round(cav, 2), auclast = round(auclast, 1)
  ) |>
  dplyr::rename(
    "Regimen" = regimen, "N" = n,
    "Cmax,ss (mg/L)" = cmax, "Tmax,ss (h)" = tmax,
    "Ctrough,ss (mg/L)" = cmin, "Ctrough 5th-95th pctile" = cmin_range,
    "Cavg,ss (mg/L)" = cav, "AUC0-tau,ss (mg*h/L)" = auclast
  ) |>
  knitr::kable(caption = "Median steady-state NCA parameters by regimen (PKNCA, final dosing interval after 40 days, 200 virtual patients per arm). Ctrough,ss is PKNCA's cmin over the interval.")
```

## Comparison against published NCA

**The source paper reports no NCA parameters.** Table 2 holds only model
parameters; there is no Cmax, Tmax, AUC or half-life anywhere in the paper or
its four supplements, and the published validation is the agreement analysis
described above. `nlmixr2lib::ncaComparisonTable()` is therefore not used here,
because there is no reference column to put beside the simulated one.

The only exposure numbers the paper does print are the two trough cut-offs the
Groningen protocol uses to trigger TDM follow-up: 1.07 mg/L for once-daily and
2.60 mg/L for twice-daily dosing (Discussion). These are decision thresholds
taken from the local protocol, not model outputs and not population
predictions, so they are recorded below as orientation rather than gated: an
adherent typical patient is expected to sit above them, which is the direction
the model gives.

```{r cutoff-orientation}
typical_ss <- function() {
  # One typical patient per regimen at the Table 1 median covariates, no IIV.
  med <- tibble::tibble(SEXF = 0, HT = 173, WT = 72, CRCL = 95)
  out <- lapply(seq_len(nrow(regimens)), function(i) {
    reg <- regimens[i, ]
    last_dose <- reg$ii * (reg$n_dose - 1L)
    ev <- dplyr::bind_rows(
      med |> tidyr::crossing(time = reg$ii * seq(0L, reg$n_dose - 1L)) |>
        dplyr::mutate(id = 1L, amt = reg$amt, evid = 1L, cmt = "depot"),
      med |> tidyr::crossing(time = seq(last_dose, last_dose + reg$ii, length.out = 97)) |>
        dplyr::mutate(id = 1L, amt = NA_real_, evid = 0L, cmt = "central")
    ) |>
      dplyr::arrange(time, evid)
    s <- rxode2::rxSolve(typ, ev, omega = NA, returnType = "data.frame")
    s <- s[!is.na(s$Cc), ]
    tibble::tibble(
      regimen = reg$regimen, cmax = max(s$Cc), ctrough = min(s$Cc),
      cutoff = if (reg$ii == 24) 1.07 else 2.60
    )
  })
  dplyr::bind_rows(out)
}
typ_ss <- typical_ss()

typ_ss |>
  dplyr::transmute(
    regimen,
    `Typical Cmax,ss (mg/L)` = round(cmax, 2),
    `Typical Ctrough,ss (mg/L)` = round(ctrough, 2),
    `Protocol follow-up cut-off (mg/L)` = cutoff,
    `Above cut-off` = ctrough > cutoff
  ) |>
  dplyr::rename("Regimen" = regimen) |>
  knitr::kable(caption = "Typical-value steady-state troughs at the Table 1 median male covariates, beside the follow-up cut-offs quoted in the Discussion. Orientation only -- the cut-offs are protocol decision thresholds, not published model predictions.")

# Gated only on being physically sensible and above the 0.2 mg/L assay LLOQ:
# this is a deterministic typical-value quantity, so the bound is tight.
stopifnot(
  all(typ_ss$ctrough > 0.2), # the paper's lower limit of quantification
  all(typ_ss$cmax > typ_ss$ctrough),
  all(typ_ss$cmax < 20) # a mis-scaled volume or dose would blow this
)
```

# Assumptions and deviations

## Interpretation of the reported "Mean (SD)" pairs

MWPharm's `KinPop` reports each population parameter as a mean with a separate
interindividual SD in the parameter's own units (Table 2; Supplement 1 and
Supplement 3 print the same `value (SD)` shape), while the Methods state that
"pharmacokinetic parameters were assumed to be log-normally distributed". The
packaged model follows this library's standing convention for that combination:
the printed mean is carried as the typical value and the printed SD is converted
to a log-scale variance by `omega^2 = log(1 + CV^2)` with `CV = SD / mean`,
giving 0.355749 for `CLm` (CV 65.4%) and 0.912284 for `Vd` (CV 122.1%).

The alternative reading is that the printed mean is the *arithmetic* mean of the
log-normal, in which case the median would be `mean / sqrt(1 + CV^2)` -- 7.93
rather than 9.47 L/h/70 kg for `CLm`, and 1.35 rather than 2.13 L/kg LBMc for
`Vd`. The paper does not say which it is, and MWPharm's convention is not
documented in it. Two observations in each direction, recorded so a reader can
judge:

* *for the arithmetic-mean reading*: it makes Models 1 and 2 of Table 2 agree on
  a median `CLm` (7.63 vs 7.93 L/h/70 kg) where the median reading has them
  18% apart, and it puts the implied median `Vd` (1.35) within 5% of the
  Arab-Alameddine literature prior of 1.42 L/kg LBMc that Supplement 1 carries.
* *for the convention used here*: it is what the rest of this library does with a
  printed mean, it keeps the packaged typical values byte-identical to Table 2
  so the source trace is checkable by eye, and it gives a typical terminal
  half-life of 12.8 h, closer to darunavir's reported value than the 10.5 h the
  alternative gives.

A user who wants the other reading can divide each typical value by
`sqrt(1 + CV^2)` without touching the variances.

## The `fd = 5` domain bound

Documented at length in its own section above, and repeated here because it is
the deviation most likely to bite. `fd = 5` extrapolates the lean-body-mass
correction past total body weight, so `LBMc = 5 * WT - 4 * LBM` and the volume
of distribution is **negative** for any subject with `WT <= 0.8 * LBM`. That
region is reachable inside the covariate ranges the paper itself reports for its
own cohort. This is a property of the published model, not of the
transcription; the virtual cohort here is restricted to the positive-`LBMc`
region and the number of draws that restriction removed is printed.

A related point: the Methods sentence "V1 is the volume of distribution (in
liters per 70 kg LBMc)" is inconsistent with Table 2's own unit label
`L/kgLBMc`, with the parallel sentence describing `V2` as "in liters per kg
LBMc", and with dimensional sense -- the per-70-kg reading would give a typical
volume of 2.6 L. The per-kg reading of Table 2 is used.

## `fr = 0` makes the `CRCL` column inert

The published clearance equation is `CL = CLm * (WT / 70) + fr * CLcr`, and
`fr` is fixed at 0 in the final model, so creatinine clearance has no effect on
any prediction. The arm is nevertheless retained in the model file, with `fr`
encoded as `e_crcl_cl_renal <- fixed(0)`, because Supplement 4 shows it was
tested at the SPC value of 0.12 and rejected on AIC (1611.08 vs 1584.89) rather
than never considered. A user can restore it by overriding
`e_crcl_cl_renal`; the mutation control above verifies that doing so changes
clearance by exactly `fr * CLcr`. The consequence for simulation is that a
`CRCL` column must be supplied even though it does nothing.

CKD-EPI needs age and sex as well as serum creatinine, and the paper does not
state whether a race coefficient was applied, so a user deriving the column
from raw creatinine will have to choose a CKD-EPI variant.

## Other assumptions

* **Race and ethnicity** are not reported in Table 1 and are not covariates, so
  the virtual cohort does not model them.
* **Height and weight distributions.** Table 1 gives medians and ranges but no
  SDs or joint distribution. The cohort samples sex-appropriate heights and a
  BMI centred on the reported median of 24.6 kg/m^2, then derives weight, so
  that height and weight stay physiologically paired -- which matters here
  because the `fd = 5` bound is a joint constraint on the two.
* **Covariates held constant.** The source treats weight and serum creatinine as
  "continuously changing covariates" (Discussion) -- that flexibility is the
  reason the authors chose ITSB over mixed-effects modelling. The cohort here
  holds them fixed per subject, which is the right choice for a steady-state
  exposure illustration but does not exercise the time-varying path.
* **Residual error combination.** The assay SD is reported as the linear sum
  `0.2 + 0.05 * C`, which is nlmixr2's `combined1()` form rather than the
  default quadrature combination of `add()` and `prop()`. Both terms are fixed,
  because they describe the assay rather than a fitted quantity.
* **Only the final model is packaged.** Table 2's Model 1 (AIC 945.31, `CLm`
  Bayesian only) is a rejected development submodel: its volume was held at the
  literature value, which "implies that the volume of distribution ... is the
  same for each patient, which does not seem logical" (Results), and its
  external validation was markedly worse (r^2 = 0.85 against 0.97; 92.3% of
  predictions within the limits of agreement against 98.2%). Per the
  base-versus-final policy the library packages the final model only. The
  two-compartment model explored in Supplement 3 is likewise not packaged; the
  paper rejected it.
* **No published NCA and no reproducible published figure.** Discussed in the
  PKNCA section. Every numeric gate in this vignette is therefore either an
  algebraic identity of the published equations or a mutation control on one of
  them, not a comparison against a published exposure number -- because the
  paper contains none.
* **Ka is not identifiable in these data.** It is fixed at the Arab-Alameddine
  literature value of 1.04 1/h because "the low number of darunavir samples
  drawn in the absorption phase; 0-4 hours after drug intake" made estimation
  impossible (Discussion). Simulated `Tmax` and the shape of the absorption
  phase are therefore inherited from that external source, not fitted here, and
  should not be treated as validated by this paper.
* **Unmodelled mechanisms the authors name.** Food intake, pharmacogenomics and
  binding to alpha-1-acid glycoprotein are all absent from the model and are
  listed in the Discussion as candidate explanations for the residual
  overestimation at low concentrations. Ritonavir is present in every regimen as
  a fixed-dose booster and is not modelled as a covariate, so the packaged model
  describes boosted darunavir only.

```{r session}
sessionInfo()
```
