---
title: "Postpartum maternal physiology and milk composition (Deferm 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Postpartum maternal physiology and milk composition (Deferm 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Deferm_2025_lactation_physiology"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Article: <https://doi.org/10.3389/fphar.2025.1517069>\n")
cat("* PubMed Central: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC11830814/>\n")
```

Deferm 2025 is not a drug model. It is the **system-parameter layer** that a
lactation physiologically-based pharmacokinetic (PBPK) model sits on top of: a
meta-analysis of 230 studies that produces continuous mathematical functions of
postpartum age for the 13 maternal and breast-milk properties that govern how
much drug reaches a breastfed infant. Published lactation PBPK models routinely
hold these constant; the paper's point is that they are not constant, and that
they return to pre-pregnancy values at very different rates.

The packaged model therefore has no drug, no dose, no compartments and no ODEs.
Every output is an algebraic function of the `rxode2` time variable, which the
model interprets as **postpartum age in months** (the paper's symbol `PpT`).
Solving over `time = 0` to `12` traces the whole first postpartum year.

```{r description, echo = FALSE, results = "asis"}
cat("> ", ui$description, "\n", sep = "")
```

# Population

The pooled database covers **20,801 postpartum women** contributing **36,689
data points** across **230 studies**, from immediately after childbirth to 12
months postpartum. The women had a weighted mean age of 28.59 years (range 20.8
to 40), weighted mean weight 63.93 kg (range 45 to 100.40) and weighted mean
height 163.78 cm (range 149 to 173.50).

Inclusion required healthy breastfeeding women aged 18 to 45 after an
uncomplicated, full-term pregnancy who took no medication during or after
pregnancy (Methods 2.3). Studies focused on preterm infants were excluded, and
mixed cohorts were used only where full-term cases were at least 90% of the
study.

Two features of the data are important when reading the validation below.
First, the data are heavily front-loaded: 60% of all data points fall within
the first postpartum month, and only 251 data points exist beyond 7 months.
Second, for four parameters (haematocrit, alpha-1-acid glycoprotein, human
serum albumin, plasma volume) no observed data extended to 12 months, so the
authors anchored the tail with a single value simulated from 1,000 healthy
non-pregnant women in the Simcyp Simulator V23R2 and assumed it represented the
12-month level. Those tails are an assumption of return-to-baseline, not an
observation.

Parity, ethnicity, delivery type, complementary-feeding practice and
foremilk-versus-hindmilk sampling were all screened and all discarded for
inconsistent reporting (Results 3.1 and Discussion); the first two are recorded
in the model's `covariatesDataExcluded` metadata.

```{r population, echo = FALSE}
pop <- ui$population
tibble::tibble(
  Field = names(pop),
  Value = vapply(pop, function(x) paste(as.character(x), collapse = "; "), character(1))
) |>
  dplyr::filter(Field %in% c("species", "n_subjects", "n_studies", "n_datapoints",
                             "age_range", "weight_range", "height_range",
                             "postpartum_range")) |>
  knitr::kable(caption = "Population metadata carried in the model file.")
```

# Source trace

Every value is a published point estimate of a function the authors fitted by
weighted least squares in Excel Solver, with each data point weighted by study
n (Methods 2.4). **The paper reports no standard error, RSE or confidence
interval for any coefficient**, so every entry is encoded with `fixed()`.

| Output (units) | Source equation | Functional form |
|---|---|---|
| `milk_volume` (L/day) | Eq. 5 (0 to 6 mo), Eq. 6 (6 to 12 mo) | sigmoidal rise, then mono-exponential decline |
| `milk_ph` (unitless) | Eq. 7 | sum of two exponentials |
| `milk_fat` (g/dL) | Eq. 8 | baseline x second-order polynomial |
| `milk_protein` (% w/v) | Eq. 9 | exponential decline to a plateau |
| `milk_water` (% w/w) | Section 3.2.5 | constant (no age trend detected) |
| `milk_intake` (L/kg/day) | Eq. 10 (0 to 1 mo), Eq. 11 (1 to 12 mo) | sigmoidal rise, then mono-exponential decline |
| `hct` (%) | Eq. 12 | sigmoidal recovery |
| `agp` (g/L) | Eq. 13 (0 to 1 mo), Eq. 14 (1 to 12 mo) | double exponential, then linear |
| `hsa` (g/L) | Eq. 15 | logistic recovery |
| `breast_volume` (L) | Eq. 16 | linear decline |
| `plasma_volume` (L) | Eq. 17 | geometric decay to an asymptote |
| `cardiac_output` (L/h) | Eq. 18 | sum of two exponentials |
| `mgfr` (mL/min) | Eq. 19 | quartic polynomial |

The per-coefficient origin is recorded as an in-file comment beside each
`ini()` entry in `inst/modeldb/endogenous/Deferm_2025_lactation_physiology.R`.
The table below reads them straight out of the packaged model.

```{r ini-table, echo = FALSE}
ini <- ui$iniDf
ini |>
  dplyr::filter(is.na(neta1)) |>
  dplyr::transmute(
    Parameter = name,
    Value = signif(est, 6),
    Fixed = ifelse(fix, "yes", "no"),
    Label = label
  ) |>
  knitr::kable(caption = "Structural coefficients as packaged (all fixed; Deferm 2025 Eqs. 5 to 19).")
```

## Dimensional analysis

Each function maps postpartum age in months to its own output unit; no term
mixes units. The only conversions a downstream user must be aware of are
collected here.

| Quantity | Unit as packaged | Note |
|---|---|---|
| time (`ppt`) | month | Every rate constant and polynomial coefficient is per month (or per month^k). Solving on an hour or day axis silently rescales all 13 functions. |
| `milk_volume` | L/day | Daily maternal production, not per feed. |
| `milk_intake` | L/kg/day | Per kg of **infant** body weight; multiply by infant weight for L/day. |
| `milk_fat` | g/dL | Numerically equal to % w/v. Creamatocrit readings were converted with the Lucas equation before fitting (Section 3.2.3). |
| `milk_protein` | % w/v | Total protein, not true protein. |
| `milk_water` | % w/w | Mass fraction; see the mass-balance check below for the w/v vs w/w caveat. |
| `cardiac_output` | L/h | The paper reports L/h throughout; divide by 60 for the more familiar L/min (403 L/h = 6.7 L/min). |
| `mgfr` | mL/min | Absolute, not body-surface-normalised. |
| `plasma_volume`, `breast_volume` | L | `breast_volume` is *empty* breast volume: total minus the milk it holds (Section 3.6). |

The three Hill-type denominators (`milkvol_t50`, `milkintake_t50`, `hct_t50`)
are raised to the same exponent as the numerator, so they carry units of
month^hill and the ratio is dimensionless, as required. `plasmavol_decay` is a
dimensionless base raised to `ppt`.

# Simulation

Solving requires only an observation grid: there is nothing to dose.

```{r simulate}
mod <- readModelDb("Deferm_2025_lactation_physiology")

# Typical-value trajectory. The model exposes each function's typical value as
# a `tv_*` column, which is unaffected by the etas, so a single solve yields
# both the typical curve and the individual values.
grid_typ <- rxode2::et(seq(0, 12, by = 0.02))
sim_typ <- rxode2::rxSolve(mod, grid_typ, nSub = 1L, returnType = "data.frame")

# Virtual cohort of 200 postpartum women (the per-arm cap for these vignettes).
set.seed(20250203)
grid_pop <- rxode2::et(seq(0, 12, by = 0.1))
sim_pop <- rxode2::rxSolve(mod, grid_pop, nSub = 200L, returnType = "data.frame")

# An nSub-generated cohort keys subjects on `sim.id`, not `id`. Normalise so
# the downstream grouping code does not depend on which one rxode2 returns.
if (is.null(sim_pop$id)) sim_pop$id <- sim_pop$sim.id
stopifnot(!is.null(sim_pop$id), dplyr::n_distinct(sim_pop$id) == 200L)

outputs <- c("milk_volume", "milk_ph", "milk_fat", "milk_protein", "milk_water",
             "milk_intake", "hct", "agp", "hsa", "breast_volume",
             "plasma_volume", "cardiac_output", "mgfr")
out_units <- c(milk_volume = "L/day", milk_ph = "pH", milk_fat = "g/dL",
               milk_protein = "% w/v", milk_water = "% w/w",
               milk_intake = "L/kg/day", hct = "%", agp = "g/L", hsa = "g/L",
               breast_volume = "L", plasma_volume = "L",
               cardiac_output = "L/h", mgfr = "mL/min")
stopifnot(all(outputs %in% names(sim_typ)), all(outputs %in% names(sim_pop)))
```

A small helper reads the typical value of any output at an arbitrary
postpartum age, by solving the model at exactly those times.

`rxode2::et()` **sorts and de-duplicates** the times it is given, so the i-th
row of a solve is not in general the i-th requested time. Indexing the solve
positionally would silently mis-attribute every value. The helper therefore
solves on the sorted unique grid and matches each requested time back by value.

```{r helper}
typical_at <- function(times) {
  u <- sort(unique(times))
  s <- rxode2::rxSolve(mod, rxode2::et(u), nSub = 1L, returnType = "data.frame")
  stopifnot(nrow(s) == length(u))
  idx <- vapply(times, function(x) which.min(abs(s$time - x)), integer(1))
  # Fail loudly rather than return a nearest-but-wrong row.
  stopifnot(max(abs(s$time[idx] - times)) < 1e-9)
  s[idx, , drop = FALSE]
}

# Guard: the helper must respect the requested order, not the sorted order.
local({
  probe <- typical_at(c(12, 0, 2))
  stopifnot(all.equal(probe$time, c(12, 0, 2)))
})
```

# Validation 1: published summary values

The paper's Results sections quote observed weighted-mean values at named
postpartum ages. Those are the observed data the functions were fitted *to*, so
the fitted curve is expected to approximate rather than interpolate them. The
table compares each quoted value against the packaged model, flagging any
difference beyond 20%.

Ages quoted in weeks are converted at 4.348 weeks per month.

```{r checkpoints}
wk <- 1 / 4.348  # months per week

checkpoints <- tibble::tribble(
  ~output,          ~ppt,       ~when,                      ~observed, ~source,
  "milk_ph",        0,          "colostrum (t = 0)",             7.43, "Section 3.2.2",
  "milk_ph",        2 * wk,     "2 weeks (observed nadir)",      7.05, "Section 3.2.2",
  "milk_water",     6,          "any age (no trend)",            87.5, "Section 3.2.5",
  "hct",            0,          "at birth",                      31.0, "Section 3.3",
  "hct",            2 * wk,     "2 weeks",                       40.0, "Section 3.3",
  "agp",            1 * wk,     "1 week (peak)",                 1.28, "Section 3.4",
  "hsa",            1,          "1 month",                       47.0, "Section 3.5",
  "hsa",            1.58,       "1.58 months",                  45.89, "Section 3.5",
  "hsa",            2.32,       "2.32 months (paper flags)",    33.70, "Section 3.5",
  "hsa",            2.71,       "2.71 months",                  45.60, "Section 3.5",
  "breast_volume",  0.23,       "0.23 months",                   1.56, "Section 3.6",
  "breast_volume",  3.63,       "3.63 months",                   1.41, "Section 3.6",
  "breast_volume",  5.83,       "5.83 months",                   1.38, "Section 3.6",
  "breast_volume", 11.82,       "11.82 months",                  1.28, "Section 3.6",
  "plasma_volume",  0,          "at birth",                      2.35, "Section 3.7",
  "plasma_volume",  1 * wk,     "1 week",                        2.53, "Section 3.7",
  "cardiac_output", 0,          "immediately after delivery",  402.25, "Section 3.8",
  "cardiac_output", 2 * wk,     "2 weeks",                     308.15, "Section 3.8",
  "cardiac_output", 4 * wk,     "end of week 4",               302.42, "Section 3.8",
  "cardiac_output", 12,         "12 months",                   311.40, "Section 3.8",
  "mgfr",           0,          "immediate postpartum",        152.00, "Section 3.9",
  "mgfr",           2,          "2 months (nadir)",             92.00, "Section 3.9"
)

chk_sim <- typical_at(checkpoints$ppt)
```

The `tv_` column names shorten the output names, so map them explicitly rather
than by string surgery -- a silent mismatch here would make the whole table
meaningless.

```{r checkpoints-map}
tv_of <- c(milk_volume = "tv_milkvol", milk_ph = "tv_milkph",
           milk_fat = "tv_milkfat", milk_protein = "tv_milkprot",
           milk_water = "tv_milkwater", milk_intake = "tv_milkintake",
           hct = "tv_hct", agp = "tv_agp", hsa = "tv_hsa",
           breast_volume = "tv_breastvol", plasma_volume = "tv_plasmavol",
           cardiac_output = "tv_cardiacout", mgfr = "tv_mgfr")
stopifnot(setequal(names(tv_of), outputs), all(tv_of %in% names(sim_typ)))

checkpoints$model <- vapply(
  seq_len(nrow(checkpoints)),
  function(i) chk_sim[[tv_of[[checkpoints$output[i]]]]][i],
  numeric(1)
)

checkpoints <- checkpoints |>
  dplyr::mutate(
    pct_diff = 100 * (model - observed) / observed,
    flag = ifelse(abs(pct_diff) > 20, "*", "")
  )

checkpoints |>
  dplyr::transmute(
    "Parameter" = paste0(output, " (", out_units[output], ")"),
    "Postpartum age" = when,
    "Observed (paper)" = signif(observed, 5),
    "Model (typical)" = signif(model, 5),
    "Difference (%)" = paste0(sprintf("%+.1f", pct_diff), flag),
    "Source" = source
  ) |>
  knitr::kable(
    align = c("l", "l", "r", "r", "r", "l"),
    caption = "Packaged model vs the observed summary values quoted in Deferm 2025. * differs by more than 20%."
  )
```

```{r checkpoints-assert}
n_flagged <- sum(checkpoints$flag == "*")
cat("Checkpoints compared:", nrow(checkpoints), "\n")
cat("Flagged (>20% difference):", n_flagged, "\n")
cat("Median absolute difference: ",
    sprintf("%.1f%%", median(abs(checkpoints$pct_diff))), "\n", sep = "")
# The comparison must actually have rows to test (a zero-row table would
# vacuously "pass"), and the tolerances are pinned to what is actually
# achieved so that a future regression in any coefficient fails here.
stopifnot(
  nrow(checkpoints) == 22L,
  !anyNA(checkpoints$model),
  n_flagged == 1L,                                  # only the m-HSA outlier
  checkpoints$output[checkpoints$flag == "*"] == "hsa",
  median(abs(checkpoints$pct_diff)) < 3,            # achieved: 2.3%
  # Every non-outlier checkpoint agrees to within 20%.
  all(abs(checkpoints$pct_diff[checkpoints$flag != "*"]) < 20)
)
```

Reading the residual differences:

* The **only** flagged row is human serum albumin at 2.32 months, where the
  paper's own pooled observation is 33.70 g/L against a model value of 44.85.
  Deferm 2025 Section 3.5 singles out exactly this stretch as anomalous --
  45.89 g/L at 1.58 months, 33.70 at 2.32, then back to 45.60 at 2.71 -- and
  states that "despite this variability, the overall trend indicated that after
  approximately 1 month postpartum ... m-HSA levels remained stable up to 12
  months". The fitted logistic deliberately does not chase the 2.32-month
  point, and neither does this model. No parameter was tuned.
* Human serum albumin at 1 month (41.5 vs 47.0, -11.7%) and alpha-1-acid
  glycoprotein at its peak (1.13 vs 1.28, -11.4%) reflect the same weighted
  fit falling below the highest observed means. Figure 9B of the paper shows
  exactly this for alpha-1-acid glycoprotein: its simulated-mean line peaks
  near 1.15 g/L while the observed points reach 1.28.
* Plasma volume at birth (2.78 vs 2.35, +18.1%) is structural: the observed
  data dip at delivery (2.35 L), rebound to 3.08 L at 0.71 weeks and fall back
  to 2.53 L at 1 week, and the authors chose a monotone geometric decay that
  cannot reproduce a dip-and-rebound. The function threads between them
  instead, landing 8.2% above the 1-week point.
* Every checkpoint not discussed above agrees to within 5.7%, the median
  absolute difference across all 22 is 2.3%, and the two GFR anchors -- the
  values the paper leads its abstract with -- agree to within 0.6%.

# Validation 2: piecewise continuity

Three parameters are defined by two functions joined at a knot. The paper does
not state that the branches were constrained to be continuous, so agreement at
the knot is an independent check that both branches were transcribed correctly:
a mis-read coefficient in either branch would open a visible step.

```{r continuity}
eps <- 1e-6
knots <- tibble::tribble(
  ~output,       ~knot, ~equations,
  "milk_volume",     6, "Eq. 5 / Eq. 6",
  "milk_intake",     1, "Eq. 10 / Eq. 11",
  "agp",             1, "Eq. 13 / Eq. 14"
)

cont <- knots |>
  dplyr::rowwise() |>
  dplyr::mutate(
    below = typical_at(c(knot - eps, knot + eps))[[tv_of[[output]]]][1],
    above = typical_at(c(knot - eps, knot + eps))[[tv_of[[output]]]][2],
    rel_step_pct = 100 * abs(above - below) / below
  ) |>
  dplyr::ungroup()

cont |>
  dplyr::transmute(
    "Parameter" = output,
    "Knot (months)" = knot,
    "Equations" = equations,
    "Left limit" = signif(below, 5),
    "Right limit" = signif(above, 5),
    "Step (%)" = sprintf("%.2f", rel_step_pct)
  ) |>
  knitr::kable(align = c("l", "r", "l", "r", "r", "r"),
               caption = "Continuity of the packaged model at the three piecewise knots.")

stopifnot(nrow(cont) == 3L, all(cont$rel_step_pct < 1))
```

All three branches meet to better than 1%. This is the check that catches the
alpha-1-acid glycoprotein sign error discussed below: with Equation 14's slope
as printed the step at the 1-month knot is 4.4%, roughly seven times larger
than with the corrected sign.

```{r continuity-agp-printed}
agp_eq13 <- function(t) exp(-1.277 * t) - exp(-6.749 * t) + 0.6
agp_eq14 <- function(t, slope) slope * t + 0.90
tibble::tibble(
  "Eq. 14 slope" = c("-0.016 (as packaged)", "+0.016 (as printed)"),
  "Eq. 13 at 1 month" = signif(agp_eq13(1), 4),
  "Eq. 14 at 1 month" = signif(c(agp_eq14(1, -0.016), agp_eq14(1, 0.016)), 4),
  "Step (%)" = sprintf(
    "%.1f",
    100 * abs(c(agp_eq14(1, -0.016), agp_eq14(1, 0.016)) - agp_eq13(1)) / agp_eq13(1)
  ),
  "Value at 12 months" = signif(c(agp_eq14(12, -0.016), agp_eq14(12, 0.016)), 4)
) |>
  knitr::kable(align = c("l", "r", "r", "r", "r"),
               caption = "Equation 14 slope: packaged (negative) vs printed (positive).")
```

# Validation 3: milk macronutrient mass balance

Water, fat and total protein are fitted independently from three different
literature sets, so their sum is unconstrained. The Discussion states that
human milk is "approximately 88% water, 7% carbohydrates, 1% protein, and 4%
fat" (Kunz 1999). Treating the residue as carbohydrate gives a check that no
single function can pass on its own.

The comparison mixes % w/w (water) with g/dL, which equals % w/v; at a milk
density near 1.03 g/mL the two differ by about 3%, well inside the width of
the check.

```{r mass-balance}
mb <- sim_typ |>
  dplyr::transmute(
    ppt = time,
    water = tv_milkwater,
    fat = tv_milkfat,
    protein = tv_milkprot,
    residue = 100 - water - fat - protein
  )

mb |>
  dplyr::filter(ppt %in% c(0, 0.5, 1, 3, 6, 12)) |>
  dplyr::transmute(
    "Postpartum age (months)" = ppt,
    "Water (% w/w)" = round(water, 2),
    "Fat (g/dL)" = round(fat, 2),
    "Protein (% w/v)" = round(protein, 2),
    "Implied carbohydrate (%)" = round(residue, 2)
  ) |>
  knitr::kable(align = "r",
               caption = "Milk macronutrient balance across the postpartum year.")

cat("Implied carbohydrate range over 0-12 months: ",
    sprintf("%.2f%% to %.2f%%", min(mb$residue), max(mb$residue)), "\n", sep = "")
# Kunz 1999 (cited in the Discussion) gives ~7% carbohydrate; human milk
# lactose is reported between about 6 and 8 g/dL across the literature.
stopifnot(nrow(mb) > 0, all(mb$residue > 5), all(mb$residue < 8.5))
```

The residue stays between 6.46% and 7.54% across the whole postpartum year,
bracketing the 7% carbohydrate figure the paper cites. Three functions fitted
from three disjoint literature sets closing to within about half a percentage
point of an independently reported constant is meaningful evidence that all
three were transcribed on the right scale -- a factor-of-ten slip on any one of them (for
example reading milk fat as 36.9 rather than 3.69 g/dL) would drive the residue
far negative.

# Validation 4: direction and shape

The Results and Discussion make explicit qualitative claims about the shape of
each trajectory. Each is asserted here so that a future regression in any
coefficient fails loudly.

```{r shape-checks}
tv <- sim_typ
t_ <- tv$time
at <- function(x) which.min(abs(t_ - x))  # nearest grid index, float-safe
peak_agp_i <- which.max(tv$tv_agp)

shape <- tibble::tribble(
  ~claim, ~source, ~holds,
  "Milk volume rises to a plateau, holds to ~6 months, then declines",
  "Section 3.2.1",
  tv$tv_milkvol[at(1)] > 0.99 * max(tv$tv_milkvol) &&
    tv$tv_milkvol[at(12)] < 0.6 * max(tv$tv_milkvol),

  "Milk pH falls sharply then rises slowly; colostrum pH exceeds mature-milk pH",
  "Section 3.2.2 / Figure 3",
  tv$tv_milkph[1] > max(tv$tv_milkph[t_ > 0.4]) &&
    tv$tv_milkph[at(12)] > min(tv$tv_milkph),

  "Milk fat increases monotonically over the first year",
  "Section 3.2.3",
  all(diff(tv$tv_milkfat) > 0),

  # "Declines and stabilises" = non-increasing everywhere, strictly
  # decreasing while the exponential term is still alive. Beyond ~1.5
  # months exp(-5.058 * ppt) has underflowed against the 1.219 plateau,
  # so consecutive values are exactly equal and a strict `< 0` test on
  # the whole grid would fail on a correct model.
  "Milk total protein declines and stabilises by ~1.5 months",
  "Section 3.2.4",
  all(diff(tv$tv_milkprot) <= 0) &&
    all(diff(tv$tv_milkprot[t_ <= 1.5]) < 0) &&
    abs(tv$tv_milkprot[at(1.5)] - tv$tv_milkprot[at(12)]) < 0.01,

  "Milk water content is constant",
  "Section 3.2.5",
  diff(range(tv$tv_milkwater)) < 1e-9,

  "Daily milk intake peaks between 0.5 and 1 month, then declines",
  "Section 3.2.6",
  t_[which.max(tv$tv_milkintake)] >= 0.5 && t_[which.max(tv$tv_milkintake)] <= 1.0,

  "Haematocrit rises to a plateau within ~2 weeks and stays there",
  "Section 3.3",
  tv$tv_hct[at(2 / 4.348)] > 0.99 * max(tv$tv_hct) &&
    all(diff(tv$tv_hct) > -1e-9),

  "Alpha-1-acid glycoprotein peaks in the first month then DECLINES to pre-pregnancy",
  "Section 3.4 / Figure 9A",
  t_[peak_agp_i] < 1 && tv$tv_agp[at(12)] < tv$tv_agp[at(1)],

  "Human serum albumin rises to pre-pregnancy levels by ~1 month then is stable",
  "Section 3.5",
  tv$tv_hsa[at(1)] > 0.9 * max(tv$tv_hsa) &&
    diff(range(tv$tv_hsa[t_ >= 3])) < 0.01,

  "Empty breast volume declines linearly",
  "Section 3.6",
  all(abs(diff(diff(tv$tv_breastvol))) < 1e-9) && all(diff(tv$tv_breastvol) < 0),

  "Plasma volume decreases to pre-pregnancy levels by ~2 months",
  "Section 3.7",
  all(diff(tv$tv_plasmavol) < 0) && abs(tv$tv_plasmavol[at(2)] - 2.67) < 0.01,

  "Cardiac output falls steeply then is stable near pre-pregnancy",
  "Section 3.8",
  all(diff(tv$tv_cardiacout) < 0) &&
    diff(range(tv$tv_cardiacout[t_ >= 2])) < 10,

  # The paper's OBSERVED nadir is at 2 months (92 mL/min); the fitted
  # quartic bottoms slightly later, at 2.66 months. Validation 1 already
  # pins the value at 2 months to the observed 92 within 0.4%, so the
  # shape claim here is only that there is an interior minimum in the
  # second-to-third month followed by recovery.
  "GFR falls to an interior nadir in months 2 to 3 then recovers",
  "Section 3.9",
  t_[which.min(tv$tv_mgfr)] > 1.5 && t_[which.min(tv$tv_mgfr)] < 3.5 &&
    tv$tv_mgfr[at(7)] > min(tv$tv_mgfr) &&
    abs(tv$tv_mgfr[at(2)] - 92) / 92 < 0.05
)

shape |>
  dplyr::transmute(
    "Published claim" = claim,
    "Source" = source,
    "Model" = ifelse(holds, "holds", "FAILS")
  ) |>
  knitr::kable(caption = "Qualitative shape claims from Deferm 2025 checked against the packaged model.")

stopifnot(nrow(shape) == 13L, all(shape$holds))
```

# Validation 5: between-subject variability

The only variability the paper reports is a constant coefficient of variation
per parameter, applied to recover the spread of the observed data (Methods 2.4
and the closing sentence of each Results subsection). The model encodes each as
an independent lognormal eta with `omega^2 = log(1 + CV^2)`. Recovering the
stated CV from a simulated cohort confirms both the conversion and that the eta
is attached to the right output.

```{r variability}
stated_cv <- c(milk_volume = 0.33, milk_ph = 0.03, milk_fat = 0.37,
               milk_protein = 0.23, milk_water = 0.015, milk_intake = 0.25,
               hct = 0.08, agp = 0.24, hsa = 0.10, breast_volume = 0.10,
               plasma_volume = 0.13, cardiac_output = 0.16, mgfr = 0.36)

at_3mo <- sim_pop |> dplyr::filter(abs(time - 3) < 1e-8)
stopifnot(nrow(at_3mo) == 200L)

cv_tab <- tibble::tibble(
  output = outputs,
  stated = stated_cv[outputs],
  empirical = vapply(outputs, function(o) {
    x <- at_3mo[[o]]
    stats::sd(x) / mean(x)
  }, numeric(1))
) |>
  dplyr::mutate(pct_diff = 100 * (empirical - stated) / stated)

cv_tab |>
  dplyr::transmute(
    "Parameter" = output,
    "Stated CV (paper)" = sprintf("%.1f%%", 100 * stated),
    "Empirical CV (n = 200)" = sprintf("%.1f%%", 100 * empirical),
    "Difference (%)" = sprintf("%+.1f", pct_diff)
  ) |>
  knitr::kable(align = c("l", "r", "r", "r"),
               caption = "Stated vs recovered coefficients of variation at 3 months postpartum (200 simulated women).")

stopifnot(nrow(cv_tab) == 13L, all(abs(cv_tab$pct_diff) < 20))
```

Every recovered CV is within 20% of its stated value, the largest deviation
being 10%. That residual scatter is the sampling error of a 200-subject cohort:
the standard error of a CV estimate at n = 200 is roughly 5% relative, so a 10%
deviation is about two standard errors.

# Replicate published figures

The panels below reproduce Figures 2 to 14 of Deferm 2025: the typical-value
trajectory (solid line) with the +/- 1 SD band from the 200-woman virtual
cohort (shaded), and the observed summary values quoted in the Results text
(points) where the paper gives them.

```{r figures, fig.width = 9, fig.height = 10}
band <- sim_pop |>
  dplyr::select(id, time, dplyr::all_of(outputs)) |>
  tidyr::pivot_longer(dplyr::all_of(outputs), names_to = "output", values_to = "value") |>
  dplyr::group_by(output, time) |>
  dplyr::summarise(
    lo = mean(value) - stats::sd(value),
    hi = mean(value) + stats::sd(value),
    .groups = "drop"
  )

line <- sim_typ |>
  dplyr::select(time, dplyr::all_of(unname(tv_of))) |>
  tidyr::pivot_longer(-time, names_to = "tvcol", values_to = "value") |>
  dplyr::mutate(output = names(tv_of)[match(tvcol, tv_of)])
stopifnot(!anyNA(line$output))

lab <- paste0(outputs, " (", out_units[outputs], ")")
names(lab) <- outputs

ggplot(band, aes(time)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey70", alpha = 0.45) +
  geom_line(data = line, aes(y = value), linewidth = 0.7, colour = "firebrick") +
  geom_point(
    data = checkpoints |> dplyr::rename(time = ppt, value = observed),
    aes(y = value), shape = 21, fill = "white", colour = "black", size = 1.8
  ) +
  facet_wrap(~output, scales = "free_y", ncol = 3,
             labeller = ggplot2::labeller(output = lab)) +
  labs(
    x = "Postpartum age (months)",
    y = NULL,
    title = "Postpartum maternal physiology and milk composition",
    caption = paste(
      "Replicates Figures 2 to 14 of Deferm 2025. Line = typical value;",
      "band = +/- 1 SD across 200 simulated women;",
      "points = observed summary values quoted in the Results text."
    )
  ) +
  theme_bw(base_size = 9) +
  theme(plot.caption = element_text(hjust = 0))
```

The early postpartum period is where almost all the data live and where the
functions move fastest, so the first month is shown on its own below --
matching panel (B) of the paper's Figures 2, 3, 5, 7, 8, 9, 10 and 13.

```{r figures-early, fig.width = 9, fig.height = 6}
early <- c("milk_volume", "milk_ph", "milk_protein", "milk_intake",
           "hct", "agp", "hsa", "cardiac_output")

ggplot(dplyr::filter(band, output %in% early, time <= 2), aes(time)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey70", alpha = 0.45) +
  geom_line(data = dplyr::filter(line, output %in% early, time <= 2),
            aes(y = value), linewidth = 0.7, colour = "firebrick") +
  geom_point(
    data = checkpoints |>
      dplyr::rename(time = ppt, value = observed) |>
      dplyr::filter(output %in% early, time <= 2),
    aes(y = value), shape = 21, fill = "white", colour = "black", size = 1.8
  ) +
  facet_wrap(~output, scales = "free_y", ncol = 4,
             labeller = ggplot2::labeller(output = lab)) +
  labs(x = "Postpartum age (months)", y = NULL,
       title = "First two months postpartum",
       caption = "Replicates panel (B) of Figures 2, 3, 5, 7, 8, 9, 10 and 13 of Deferm 2025.") +
  theme_bw(base_size = 9) +
  theme(plot.caption = element_text(hjust = 0))
```

# Using this model with a lactation PBPK model

The intended use is as a covariate layer. A drug model that needs, say, milk
pH and milk fat to compute a milk-to-plasma ratio, or maternal GFR to scale
renal clearance, can read them off a solve of this model at the postpartum ages
of interest.

```{r usage-example}
# Physiology a lactation PBPK model would see at 3 days, 1 month and 6 months.
usage <- typical_at(c(3 / 30.44, 1, 6))
usage |>
  dplyr::transmute(
    "Postpartum age (months)" = round(time, 3),
    "Milk pH" = round(tv_milkph, 3),
    "Milk fat (g/dL)" = round(tv_milkfat, 3),
    "Milk volume (L/day)" = round(tv_milkvol, 3),
    "Infant intake (L/kg/day)" = round(tv_milkintake, 4),
    "Maternal AAG (g/L)" = round(tv_agp, 3),
    "Maternal albumin (g/L)" = round(tv_hsa, 2),
    "Maternal GFR (mL/min)" = round(tv_mgfr, 1)
  ) |>
  knitr::kable(align = "r",
               caption = "Typical maternal and milk physiology at three postpartum ages.")
```

The pH shift alone is consequential: the model puts colostrum at pH 7.6 and
mature milk near 7.2, and the paper's own fluoxetine example (Pansari 2022)
could only match observed milk-to-plasma ratios in both colostrum and mature
milk once that shift was applied.

# Assumptions and deviations

* **Equation 14 sign correction (the one substantive deviation).** Equation 14
  is printed as `m-AGP (g/L) = 0.016 x PpT + 0.90`, a positive slope that rises
  from 0.92 g/L at 1 month to 1.09 g/L at 12 months. The packaged model uses
  `-0.016`. The printed sign is contradicted by the paper in three independent
  ways: (1) Section 3.4 states that after the first month "m-AGP levels were
  assumed to decline linearly ... returning to pre-pregnancy levels by 12
  months postpartum"; (2) Figure 9A shows the simulated mean line falling from
  about 0.88 g/L at 1 month to about 0.71 g/L at 12 months, which `-0.016`
  reproduces to two decimal places and `+0.016` does not; and (3) the negative
  slope joins Equation 13 at the 1-month knot to within 0.7% while the printed
  positive slope opens a 4.4% step (see Validation 2). No erratum for this
  article was found at the time of extraction. The deviation is flagged in the
  model file at both `agp_slope` and the `model()` block.
* **Distribution family for the CVs.** The paper states a constant CV per
  parameter but does not name the distribution. The model applies each CV as an
  independent lognormal multiplicative eta with `omega^2 = log(1 + CV^2)`,
  which matches the Simcyp virtual-population convention the figures were
  generated under and keeps every parameter positive. A normal (additive) CV
  would put non-physiological negative values in the tails of the wider
  parameters (milk fat CV 37%, GFR CV 36%).
* **Etas are independent.** No correlations between parameters are reported.
  In reality several are strongly linked -- the paper's own Discussion notes
  that falling plasma volume is what concentrates red cells and raises
  haematocrit -- so a cohort simulated from this model will understate the
  joint structure even though each marginal CV is correct.
* **Validity window is 0 to 12 months.** All functions were fitted over that
  range and none should be evaluated outside it. The quartic GFR polynomial
  (Equation 19) in particular diverges rapidly: it is well behaved to 12 months
  but exceeds 1,500 mL/min by 20 months.
* **Time axis is months.** The `rxode2` time variable is postpartum age in
  months. Every rate constant is per month; solving on an hour or day axis
  silently rescales all 13 functions.
* **Data beyond 7 months are sparse, and four 12-month anchors are simulated,
  not observed.** For haematocrit, alpha-1-acid glycoprotein, human serum
  albumin and plasma volume the 12-month value is a Simcyp-simulated
  non-pregnant female value that the authors assumed represents 12 months
  postpartum. The late tails of those four functions encode an assumption of
  return-to-baseline.
* **No covariates.** Parity, ethnicity, delivery type, complementary feeding
  and foremilk-versus-hindmilk sampling were screened and discarded for
  inconsistent reporting; the first two are recorded in
  `covariatesDataExcluded`. Notably this means the wide early plasma-volume and
  cardiac-output spread -- which the Discussion attributes largely to caesarean
  versus vaginal delivery and to parity -- is carried as unexplained
  variability rather than as a covariate effect.
* **Piecewise branches use indicator multiplication.** `(ppt < k) * A + (ppt >= k) * B`
  rather than `ifelse`, which is the `rxode2` idiom and evaluates both branches;
  both are finite everywhere on 0 to 12 months, so this is safe here.
* **Supplementary Table 1 was not used.** It lists the 230 contributing studies;
  the model extracts only the fitted functions from the main text, which is what
  the paper offers for integration into a PBPK model.
* **Milk composition units.** The mass-balance check mixes % w/w (water) with
  g/dL and % w/v (fat, protein), as the paper itself does. At a milk density
  near 1.03 g/mL the discrepancy is about 3%, which does not affect the
  conclusion.
* **No PKNCA validation.** There is no drug, no dose and no concentration-time
  profile, so NCA is not a meaningful check for this model. The validation
  above follows the endogenous / mechanistic pattern instead: published-value
  checkpoints, piecewise continuity, a mass-balance check, qualitative shape
  assertions, and recovery of the stated variability.
