---
title: "AZD7648 + olaparib xenograft PK-PD (DeJongh 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AZD7648 + olaparib xenograft PK-PD (DeJongh 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
set.seed(20250217)
```

# Models and source

This paper contributed three models, matching the three parameter tables and the
three NONMEM control streams the authors published:

| Model | Source | What it is |
|---|---|---|
| `DeJongh_2025_azd7648_mouse` | Table 1 | AZD7648 population PK in mice |
| `DeJongh_2025_olaparib_mouse` | Table 2 | Olaparib population PK in mice |
| `DeJongh_2025_azd7648_olaparib_xenograft_mouse` | Tables 1-3 | The combined PK-PD tumour-growth-inhibition model |

* Article: <https://doi.org/10.1007/s10928-025-09962-x>
* Supplement (supplementary files 1-7, including the three NONMEM control
  streams, the tumour-static-equilibrium derivation and the PK-PD dataset):
  <https://static-content.springer.com/esm/art%3A10.1007%2Fs10928-025-09962-x/MediaObjects/10928_2025_9962_MOESM4_ESM.txt>

```{r models}
azd <- rxode2::rxode(readModelDb("DeJongh_2025_azd7648_mouse"))
ola <- rxode2::rxode(readModelDb("DeJongh_2025_olaparib_mouse"))
tgi <- rxode2::rxode(readModelDb("DeJongh_2025_azd7648_olaparib_xenograft_mouse"))
```

* Citation: `r tgi$reference`

The authors fitted the two PK models on their own plasma datasets and then
carried **only their fixed effects** into the PK-PD analysis, because most PK
data came from satellite groups. The three files reproduce that structure: the
two PK models keep their inter-individual variability and residual error, and
the PK-PD model wraps both PK models with every PK parameter `fixed()`.

# Population

```{r population}
str(tgi$population, max.level = 1)
```

AZD7648 and olaparib plasma PK were pooled from eight and six mouse studies
respectively, in female SCID (C.B-17/IcrHan(R)Hsd-Prkdcscid) and female athymic
nude (Hsd:Athymic Nude-Foxn1nu) mice at least six weeks old. Blood was drawn by
tail-vein venipuncture (20 uL per sample, at most 100 uL over 24 h) and assayed
by UPLC-MS/MS against an 11-point curve from 1 to 10,000 nM (LLOQ 0.009 nM).

The tumour PK-PD parameters (Table 3) were fitted simultaneously to all eight
treatment groups of study S1822: 120 female SCID mice bearing a subcutaneous
FaDu ATM-knockout xenograft, randomised at a mean tumour volume of about
0.2 cm3 and dosed orally once daily for 28 days with vehicle, AZD7648
monotherapy, olaparib monotherapy or one of five combinations, then followed
through up to eight weeks of washout. Studies S1842 (six-week intermittent
schedules) and S1826 were held out for a priori forecast evaluation and did not
contribute to the estimates encoded here.

Amounts are molar. The supplied NONMEM dataset (supplementary file 7) pairs
each `AMT` in umol/kg with a `DOSE` in mg/kg, which fixes the conversions used
throughout this vignette:

```{r molar}
UMOL_PER_MG_AZD <- 262.8812 / 100   # AZD7648: 100 mg/kg = 262.8812 umol/kg
UMOL_PER_MG_OLA <- 229.8428 / 100   # olaparib: 100 mg/kg = 229.8428 umol/kg
c(AZD7648 = 1000 / UMOL_PER_MG_AZD, olaparib = 1000 / UMOL_PER_MG_OLA)  # g/mol
```

# Source trace

Every `ini()` entry carries an in-file comment naming its source location. The
table below is the same trace in one place.

```{r source-trace, echo = FALSE}
knitr::kable(tibble::tribble(
  ~Model, ~Quantity, ~Source,
  "azd7648",   "ka (SCID) 2.7726 1/h; ka (nude) 9.9 1/h",              "Table 1",
  "azd7648",   "CL 0.251 L/kg/h; Vmax 4.27 umol/kg/h; Km 3.73 uM",     "Table 1",
  "azd7648",   "Vc 3.45 L/kg (Vp = Vc); Q 0.932 L/kg/h",               "Table 1",
  "azd7648",   "Nude relative F1 effect -0.520",                       "Table 1",
  "azd7648",   "OMEGA block 0.0631 / -0.0291 / 0.193; prop. RE var 0.304", "Table 1",
  "azd7648",   "Disposition equations",                                "Equations 1-3 + supplementary file 5 $DES",
  "olaparib",  "ka 10 1/h; CL 1.95 L/h; Vc 1.17 L/kg; Q 0.668 L/h; Vp 2.66 L/kg", "Table 2",
  "olaparib",  "Per-study F1 0.425 / 0.174 / 0.489 / 0.598 / 0.238 / 1.48 / 4.18", "Table 2 + supplementary file 6 $THETA",
  "olaparib",  "CmD50 82.5 mg/kg/day; OMEGA Vc 0.591; prop. RE var 0.399", "Table 2",
  "olaparib",  "Disposition + interaction equations",                  "Equations 4-8 + supplementary file 6 $DES",
  "xenograft", "BL 0.115 cm3; GC 0.329 1/day; KC 2.25 cm3",            "Table 3",
  "xenograft", "AEC 1.93 mm3; G 0.637; Ktr 0.961 1/day",               "Table 3",
  "xenograft", "EC50az 8.99 uM; EColap 0.0154 1/uM",                   "Table 3",
  "xenograft", "Kto = GC (not independently identifiable)",            "Table 3 footnote c",
  "xenograft", "OMEGA 0.0951 / 0.135 / 1.83 and 0.0362; prop. RE SD 0.243", "Table 3",
  "xenograft", "Cell-state and biophase equations",                    "Equations 9-18 + supplementary file 4 $DES",
  "xenograft", "Olaparib F1 = 4.1817 in the xenograft studies",        "supplementary file 4 (`F11`)",
  "all",       "mg/kg to umol/kg conversions",                         "supplementary file 7 (AMT vs DOSE columns)"
))
```

# Part 1 - Pharmacokinetics

## Replicating Figure 1

Figure 1 shows visual predictive checks after a **single** oral dose in SCID
mice: 100 mg/kg AZD7648 alone (1a) and with 100 mg/kg olaparib (1b), and
100 mg/kg olaparib alone (1c) and with 100 mg/kg AZD7648 (1d). The single-dose
SCID olaparib arms come from study S1734, so that study's bioavailability
factor applies.

```{r pk-sim}
N_PK <- 100L   # subjects per arm

# Both drugs absorb very fast (ka 2.8 and 10 1/h), so Tmax sits below an hour.
# The grid is deliberately dense over the first two hours: a coarser grid
# under-resolves the peak and the linear trapezoid then understates AUC by
# several percent, which would break the exact identity checked below.
PK_GRID <- c(seq(0, 2, by = 0.01), seq(2.1, 8, by = 0.1), seq(8.5, 72, by = 0.5))

pk_events <- function(n, dose_umol, cmt_dose, cmt_obs, covs) {
  doses <- tibble::tibble(id = seq_len(n), time = 0, amt = dose_umol,
                          evid = 1L, cmt = cmt_dose)
  obs <- tidyr::crossing(id = seq_len(n), time = PK_GRID) |>
    mutate(amt = NA_real_, evid = 0L, cmt = cmt_obs)
  ev <- bind_rows(doses, obs) |> arrange(id, time, desc(evid))
  for (nm in names(covs)) ev[[nm]] <- covs[[nm]]
  ev
}

# --- AZD7648: identical with and without olaparib (no reverse interaction) ---
ev_azd <- pk_events(N_PK, 100 * UMOL_PER_MG_AZD, "depot", "central",
                    list(STRAIN_NUDE = 0))
sim_azd <- rxode2::rxSolve(azd, ev_azd, useLinCmt = FALSE,
                           returnType = "data.frame")
stopifnot(dplyr::n_distinct(sim_azd$id) == N_PK)

# --- Olaparib: monotherapy vs 100 mg/kg/day AZD7648 co-treatment ---
ola_covs <- function(dazd) {
  list(STRAIN_NUDE = 0, STUDY_S1143 = 0, STUDY_S1721 = 0, STUDY_S1734 = 1,
       STUDY_S1770 = 0, STUDY_S1816 = 0, DOSE_AZD7648_MGKGD = dazd)
}
sim_ola <- bind_rows(lapply(c(0, 100), function(dz) {
  ev <- pk_events(N_PK, 100 * UMOL_PER_MG_OLA, "depot", "central", ola_covs(dz))
  rxode2::rxSolve(ola, ev, useLinCmt = FALSE, returnType = "data.frame") |>
    mutate(arm = if (dz == 0) "olaparib alone" else "olaparib + AZD7648")
}))
stopifnot(dplyr::n_distinct(sim_ola$id) == N_PK)
```

```{r fig1, echo = FALSE, fig.cap = "Replicates Figure 1 of DeJongh 2025: single-dose plasma profiles in SCID mice. Ribbons are the 10-90% prediction interval, lines the median."}
band <- function(d, grp) {
  d |> group_by(across(all_of(grp)), time) |>
    summarise(lo = quantile(Cc, 0.10), md = median(Cc), hi = quantile(Cc, 0.90),
              .groups = "drop")
}
plot_dat <- bind_rows(
  band(mutate(sim_azd, arm = "AZD7648 100 mg/kg (a, b)"), "arm"),
  band(sim_ola, "arm")
)
ggplot(plot_dat, aes(time, md)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue") +
  facet_wrap(~arm, scales = "free_y") +
  coord_cartesian(xlim = c(0, 24)) +
  labs(x = "Time (h)", y = "Plasma concentration (uM)") +
  theme_bw()
```

Co-treatment raises olaparib exposure, which is the PK half of the paper's
explanation for the combination's efficacy. AZD7648 is unchanged by olaparib:
the interaction in this model runs in one direction only.

## PKNCA

```{r pknca}
nca_one <- function(sim, dose_umol, label) {
  conc <- sim |> filter(!is.na(Cc)) |> transmute(id, time, Cc, arm = label)
  dose <- conc |> distinct(id, arm) |> mutate(time = 0, amt = dose_umol)
  o <- PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(conc, Cc ~ time | arm + id, concu = "uM", timeu = "h"),
    PKNCA::PKNCAdose(dose, amt ~ time | arm + id, doseu = "umol/kg"),
    intervals = data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                           aucinf.obs = TRUE, half.life = TRUE)))
  as.data.frame(o$result)
}
nca <- bind_rows(
  nca_one(sim_azd, 100 * UMOL_PER_MG_AZD, "AZD7648 100 mg/kg"),
  nca_one(filter(sim_ola, arm == "olaparib alone"), 100 * UMOL_PER_MG_OLA,
          "olaparib 100 mg/kg"),
  nca_one(filter(sim_ola, arm == "olaparib + AZD7648"), 100 * UMOL_PER_MG_OLA,
          "olaparib 100 mg/kg + AZD7648")
)
```

```{r pknca-table, echo = FALSE}
nca |>
  group_by(arm, PPTESTCD) |>
  summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = PPTESTCD, values_from = median) |>
  rename("Arm" = arm, "Cmax (uM)" = cmax, "Tmax (h)" = tmax,
         "AUC0-inf (uM*h)" = aucinf.obs, "t1/2 (h)" = half.life) |>
  knitr::kable(digits = 3)
```

DeJongh 2025 reports no NCA table, so there is nothing to compare these against
directly. Instead the model is checked against three identities that follow
exactly from the published parameters.

### Check 1 - olaparib AUC equals F x Dose / CL

Olaparib elimination is linear, so `AUC(0-inf)` must equal `F * Dose / CL`
exactly, per subject. The one random effect is on Vc, which does not enter that
identity, so every subject must agree to solver tolerance.

```{r check-auc-cl}
F_S1734_SCID <- 0.59754; CL_OLA <- 1.9460
expected <- F_S1734_SCID * 100 * UMOL_PER_MG_OLA / CL_OLA
observed <- nca |> filter(arm == "olaparib 100 mg/kg", PPTESTCD == "aucinf.obs")
stopifnot(nrow(observed) == N_PK)
rel_err <- abs(observed$PPORRES - expected) / expected
c(expected = expected, max_rel_error = max(rel_err))
stopifnot(max(rel_err) < 0.002)
```

### Check 2 - the interaction halves clearance at CmD50

`CmD50 = 82.5 mg/kg/day` is defined as the AZD7648 daily dose that reduces
olaparib clearance by 50%. At that dose olaparib AUC must therefore be exactly
doubled.

```{r check-ddi}
auc_at <- function(dazd) {
  ev <- pk_events(50L, 100 * UMOL_PER_MG_OLA, "depot", "central", ola_covs(dazd))
  s <- rxode2::rxSolve(rxode2::zeroRe(ola), ev, omega = NA, useLinCmt = FALSE,
                       returnType = "data.frame")
  n <- nca_one(s, 100 * UMOL_PER_MG_OLA, "x")
  median(n$PPORRES[n$PPTESTCD == "aucinf.obs"])
}
ratio <- auc_at(82.538) / auc_at(0)
c(auc_ratio = ratio, expected = 2)
stopifnot(abs(ratio - 2) < 0.01)
```

### Check 3 - AZD7648 elimination is saturable

With a Michaelis-Menten term alongside linear clearance, dose-normalised
exposure must rise with dose.

```{r check-saturation}
dnauc <- vapply(c(25, 50, 100, 150), function(d) {
  ev <- pk_events(1L, d * UMOL_PER_MG_AZD, "depot", "central",
                  list(STRAIN_NUDE = 0))
  s <- rxode2::rxSolve(rxode2::zeroRe(azd), ev, omega = NA, useLinCmt = FALSE,
                       returnType = "data.frame")
  if (is.null(s$id)) s$id <- 1L
  n <- nca_one(s, d * UMOL_PER_MG_AZD, "x")
  n$PPORRES[n$PPTESTCD == "aucinf.obs"] / (d * UMOL_PER_MG_AZD)
}, numeric(1))
names(dnauc) <- paste0(c(25, 50, 100, 150), " mg/kg")
round(dnauc, 4)
stopifnot(all(diff(dnauc) > 0))
```

# Part 2 - The tumour PK-PD model

## Check 4 - the untreated steady state

Supplementary file 2 derives the untreated steady state analytically. With
`Kto = GC` and the Allee factor at its high-mass limit of 1, total tumour
volume settles at exactly the carrying capacity `KC`, and the four cell states
hold 40% / 20% / 20% / 20% of it.

```{r check-steadystate}
ev_ss <- tibble::tibble(id = 1L, time = seq(0, 365 * 24, by = 24),
                        amt = NA_real_, evid = 0L, cmt = "cycling_cells",
                        DOSE_AZD7648_MGKGD = 0)
ss <- rxode2::rxSolve(rxode2::zeroRe(tgi), ev_ss, omega = NA,
                      useLinCmt = FALSE, returnType = "data.frame") |> tail(1)
fractions <- with(ss, c(proliferating = cycling_cells, quiescent = damaged_cells1,
                        dying1 = damaged_cells2, dying2 = damaged_cells3) /
                    tumor_size)
round(fractions, 4)
stopifnot(max(abs(fractions - c(0.4, 0.2, 0.2, 0.2))) < 0.002)
```

The realised steady state sits just below `KC` because the Allee factor is
0.98, not 1, at that tumour size. Retaining it gives the exact balance
condition `(1 - Atot / (2 KC)) * AlleeEff = 1/2`:

```{r check-steadystate-exact}
AEC_CM3 <- 1.93 / 1000; G_HILL <- 0.637; KC <- 2.25
allee <- with(ss, cycling_cells^G_HILL / (cycling_cells^G_HILL + AEC_CM3^G_HILL))
balance <- (1 - ss$tumor_size / (2 * KC)) * allee
c(tumor_size = ss$tumor_size, KC_limit = KC, allee_factor = allee,
  balance = balance, expected = 0.5)
stopifnot(abs(balance - 0.5) < 1e-3)
```

## Replicating Figures 2 and 3 - study S1822

The eight treatment groups of study S1822, all dosed once daily for 28 days
(olaparib one hour after the morning AZD7648 dose). Observed medians are read
from the supplied PK-PD dataset (supplementary file 7).

```{r s1822-design}
groups <- tibble::tribble(
  ~grp, ~azd, ~ola, ~label,
  1L,   0,    0,    "1: vehicle",
  2L, 100,    0,    "2: AZD 100",
  3L, 100,  100,    "3: AZD 100 + Ola 100",
  4L, 100,   50,    "4: AZD 100 + Ola 50",
  5L,  50,   50,    "5: AZD 50 + Ola 50",
  6L,  50,  100,    "6: AZD 50 + Ola 100",
  7L,   0,  100,    "7: Ola 100",
  8L,  75,   75,    "8: AZD 75 + Ola 75")

tgi_events <- function(id, azd, ola, days = 77, by = 12) {
  obs <- tibble::tibble(time = seq(0, days * 24, by = by), amt = NA_real_,
                        evid = 0L, cmt = "cycling_cells", ii = 0, addl = 0)
  d <- NULL
  if (azd > 0) d <- bind_rows(d, tibble::tibble(time = 0, amt = azd * UMOL_PER_MG_AZD,
                    evid = 1L, cmt = "depot", ii = 24, addl = 27))
  if (ola > 0) d <- bind_rows(d, tibble::tibble(time = 1, amt = ola * UMOL_PER_MG_OLA,
                    evid = 1L, cmt = "depot_olaparib", ii = 24, addl = 27))
  bind_rows(d, obs) |> arrange(time, desc(evid)) |>
    mutate(id = id, DOSE_AZD7648_MGKGD = azd)
}

ev_typ <- purrr::map_dfr(seq_len(nrow(groups)),
                         \(i) tgi_events(groups$grp[i], groups$azd[i], groups$ola[i]))
sim_typ <- rxode2::rxSolve(rxode2::zeroRe(tgi), ev_typ, omega = NA,
                           useLinCmt = FALSE, returnType = "data.frame") |>
  left_join(groups, by = c("id" = "grp"))
stopifnot(dplyr::n_distinct(sim_typ$id) == 8L)
```

```{r observed}
observed <- tibble::tribble(
  ~grp, ~day, ~tv,
  1L,0,0.1196, 1L,7,0.4843, 1L,9,0.5682, 1L,14,0.6735, 1L,18,0.7746, 1L,21,1.5365,
  2L,0,0.1263, 2L,7,0.3608, 2L,9,0.3960, 2L,14,0.5236, 2L,18,0.7342, 2L,21,0.9571,
  3L,0,0.1263, 3L,7,0.2989, 3L,9,0.2917, 3L,14,0.2038, 3L,18,0.1781, 3L,21,0.0908,
  3L,28,0.0965, 3L,35,0.0397, 3L,42,0.0494, 3L,49,0.0486, 3L,56,0.0702, 3L,63,0.0985,
  3L,70,0.1692, 3L,77,0.3459,
  4L,0,0.1263, 4L,7,0.2833, 4L,9,0.3274, 4L,14,0.2665, 4L,18,0.2285, 4L,21,0.1716,
  4L,28,0.2333, 4L,35,0.1498, 4L,42,0.2729, 4L,49,0.6485, 4L,56,0.6459, 4L,63,0.9852,
  5L,0,0.1251, 5L,7,0.4024, 5L,9,0.4259, 5L,14,0.4416, 5L,18,0.4952, 5L,21,0.4798,
  5L,28,0.7619,
  6L,0,0.1259, 6L,7,0.2575, 6L,9,0.2652, 6L,14,0.2514, 6L,18,0.2285, 6L,21,0.2238,
  6L,28,0.3862, 6L,35,0.3138, 6L,42,0.6544, 6L,49,0.8796, 6L,56,1.0379,
  7L,0,0.1250, 7L,7,0.3742, 7L,9,0.4203, 7L,14,0.4915, 7L,18,0.5179, 7L,21,0.6579,
  7L,28,1.2893,
  8L,0,0.1254, 8L,7,0.3289, 8L,9,0.3124, 8L,14,0.2678, 8L,18,0.2227, 8L,21,0.1511,
  8L,28,0.2070, 8L,35,0.2269, 8L,42,0.4955, 8L,49,0.7412, 8L,56,1.2251, 8L,63,0.6633,
  8L,70,1.5712) |>
  left_join(groups, by = "grp")
```

```{r fig2, echo = FALSE, fig.height = 6, fig.cap = "Replicates Figures 2a and 3b of DeJongh 2025: observed median xenograft volume (points, supplementary file 7) against the typical-value prediction of the packaged model (lines). The vertical line marks the end of the 28-day treatment period."}
ggplot(sim_typ, aes(time / 24, tumor_size)) +
  geom_vline(xintercept = 28, linetype = "dashed", colour = "grey50") +
  geom_line(colour = "steelblue") +
  geom_point(data = observed, aes(day, tv), colour = "black", size = 1.3) +
  facet_wrap(~label, ncol = 2) +
  labs(x = "Time (days)", y = "Xenograft volume (cm3)") +
  theme_bw()
```

The model reproduces the qualitative pattern the paper describes: monotherapy
with either agent slows growth but never drives the tumour below baseline,
whereas every combination arm regresses below the pre-treatment volume, with
delayed re-growth after treatment stops. Agreement is closest during the
28-day treatment period. Median observed volumes in the washout phase rest on
progressively fewer animals (as few as one or two by day 63-70) because of the
non-random dropout the paper discusses in its Discussion, so the late points
should not be read as a like-for-like comparison.

```{r fig2-table, echo = FALSE}
observed |> filter(day %in% c(21, 28)) |>
  left_join(transmute(sim_typ, grp = id, day = time / 24, pred = tumor_size),
            by = c("grp", "day")) |>
  transmute(Group = label, Day = day, `Observed median (cm3)` = round(tv, 3),
            `Model typical value (cm3)` = round(pred, 3),
            `Ratio` = round(pred / tv, 2)) |>
  knitr::kable()
```

## The Allee effect and sustained regression

The paper's central structural claim is that an Allee term is needed to explain
why a minority of mice regress permanently while most re-grow. With
inter-individual variability switched on, the packaged model reproduces that
split in the highest-dose combination group.

```{r allee-cohort}
N_TGI <- 100L
ev_pop <- purrr::map_dfr(seq_len(N_TGI), \(i) tgi_events(i, 100, 100))
sim_pop <- rxode2::rxSolve(tgi, ev_pop, useLinCmt = FALSE,
                           returnType = "data.frame")
stopifnot(dplyr::n_distinct(sim_pop$id) == N_TGI)

# The typical-value chunks above pass zeroRe(tgi) into rxSolve(). That call
# mutates state shared with the model object, and a silent strip would collapse
# this cohort to 100 identical animals without raising an error. Recover the
# two random effects from the returned per-subject parameters and check them
# against the Table 3 omegas, so the failure cannot pass unnoticed.
etas <- distinct(sim_pop, id, rbase, kge)
stopifnot(
  nrow(etas) == N_TGI,
  abs(sd(log(etas$rbase)) - sqrt(0.0951)) < 0.10,   # Om_1 (BL)
  abs(sd(log(etas$kge))   - sqrt(0.0362)) < 0.10    # Om_3 (GC)
)

final <- sim_pop |> filter(time == max(time)) |>
  transmute(id, tumor_size, regressed = tumor_size < 0.05)
mean(final$regressed)
```

```{r fig3, echo = FALSE, fig.cap = "Replicates Figure 3b of DeJongh 2025: individual profiles in the 100 mg/kg AZD7648 + 100 mg/kg olaparib group. Blue traces are animals in sustained regression at the end of follow-up; the Allee term is what allows the two outcomes to coexist under identical dosing."}
ggplot(left_join(sim_pop, final |> select(id, regressed), by = "id"),
       aes(time / 24, tumor_size, group = id, colour = regressed)) +
  geom_vline(xintercept = 28, linetype = "dashed", colour = "grey50") +
  geom_line(alpha = 0.4) +
  scale_colour_manual(values = c(`FALSE` = "grey35", `TRUE` = "steelblue"),
                      name = "Sustained regression") +
  scale_y_log10() +
  labs(x = "Time (days)", y = "Xenograft volume (cm3, log scale)") +
  theme_bw() + theme(legend.position = "bottom")
```

## Replicating Figures 4 and 5 - the a priori forecast for study S1842

Study S1842 was **held out** of the fit: the model was calibrated on 28 days of
continuous dosing in S1822 and then used to forecast six weeks of intermittent
schedules. All seven treated groups received 100 mg/kg olaparib once daily
throughout; AZD7648 was given twice daily at 75 or 37.5 mg/kg under a
continuous or an on/off schedule (Table S1). Figures 4 and 5 of the paper are
built from these groups, so they are reproduced here rather than from S1822.

```{r s1842-design}
on_days <- function(sched) {
  d <- 0:41
  switch(sched,
    continuous = d,
    `14/14/14` = d[(d < 14) | (d >= 28)],
    `7/7`      = d[(d %% 14) < 7],
    `5/2`      = d[(d %%  7) < 5],
    `3/4`      = d[(d %%  7) < 3])
}
s1842 <- tibble::tribble(
  ~grp, ~azd_bid, ~sched,       ~label,
  2L,   75,       "continuous", "AZD 75 continuous",
  3L,   75,       "14/14/14",   "AZD 75 14/14/14",
  4L,   75,       "7/7",        "AZD 75 7/7",
  5L,   75,       "5/2",        "AZD 75 5/2",
  6L,   75,       "3/4",        "AZD 75 3/4",
  7L,   37.5,     "continuous", "AZD 37.5 continuous",
  8L,   37.5,     "5/2",        "AZD 37.5 5/2")

s1842_events <- function(id, azd_bid, sched) {
  onl <- on_days(sched)
  # AZD7648 twice daily (0 and 12 h) on treatment days only
  azd_d <- tidyr::crossing(day = onl, h = c(0, 12)) |>
    transmute(time = day * 24 + h, amt = azd_bid * UMOL_PER_MG_AZD,
              evid = 1L, cmt = "depot")
  # olaparib once daily throughout, 1 h after the morning AZD7648 dose
  ola_d <- tibble::tibble(time = (0:41) * 24 + 1, amt = 100 * UMOL_PER_MG_OLA,
                          evid = 1L, cmt = "depot_olaparib")
  obs <- tibble::tibble(time = seq(0, 42 * 24, by = 6), amt = NA_real_,
                        evid = 0L, cmt = "cycling_cells")
  bind_rows(azd_d, ola_d, obs) |> arrange(time, desc(evid)) |>
    mutate(id = id, .day = floor(time / 24),
           # the interaction is driven by the AZD7648 dose actually given that
           # day, so the covariate drops to 0 during an off-period
           DOSE_AZD7648_MGKGD = ifelse(.day %in% onl, 2 * azd_bid, 0)) |>
    select(-.day)
}
ev_1842 <- purrr::map_dfr(seq_len(nrow(s1842)),
                          \(i) s1842_events(s1842$grp[i], s1842$azd_bid[i], s1842$sched[i]))
sim_1842 <- rxode2::rxSolve(rxode2::zeroRe(tgi), ev_1842, omega = NA,
                            useLinCmt = FALSE, returnType = "data.frame") |>
  left_join(s1842, by = c("id" = "grp"))
stopifnot(dplyr::n_distinct(sim_1842$id) == 7L)
```

```{r fig4, echo = FALSE, fig.cap = "Replicates Figure 4 of DeJongh 2025: a priori forecasts of xenograft volume for the six-week intermittent schedules of study S1842, a dataset held out of the model fit."}
ggplot(sim_1842, aes(time / 24, tumor_size, colour = label)) +
  geom_hline(yintercept = 0.115, linetype = "dotted", colour = "grey40") +
  geom_line() +
  scale_y_log10() +
  labs(x = "Time (days)", y = "Xenograft volume (cm3, log scale)", colour = NULL) +
  theme_bw() + theme(legend.position = "bottom")
```

```{r fig5a, echo = FALSE, fig.height = 5.5, fig.cap = "Replicates Figure 5a of DeJongh 2025: fractional occupancy of each cell state over the six-week treatment period."}
sim_1842 |>
  transmute(label, day = time / 24,
            Proliferating = cycling_cells / tumor_size,
            Quiescent = damaged_cells1 / tumor_size,
            Apoptotic = (damaged_cells2 + damaged_cells3) / tumor_size) |>
  pivot_longer(-c(label, day), names_to = "state", values_to = "fraction") |>
  ggplot(aes(day, fraction, colour = state)) +
  geom_line() + facet_wrap(~label, ncol = 3) +
  labs(x = "Time (days)", y = "Fraction of tumour volume", colour = NULL) +
  theme_bw() + theme(legend.position = "bottom")
```

```{r fig5b-calc}
trapz <- function(t, y) sum(diff(t) * (head(y, -1) + tail(y, -1)) / 2)

occupancy <- sim_1842 |>
  group_by(grp = id, label) |>
  summarise(
    prol      = trapz(time, cycling_cells / tumor_size),
    quiesc    = trapz(time, damaged_cells1 / tumor_size),
    apop      = trapz(time, (damaged_cells2 + damaged_cells3) / tumor_size),
    max_apop  = max((damaged_cells2 + damaged_cells3) / tumor_size),
    prol_depl = 1 - min(cycling_cells / tumor_size),
    # Figure 5b efficacy: fold change 24 h after the last dose vs baseline
    fold      = tumor_size[which.max(time)] / first(tumor_size),
    .groups   = "drop") |>
  mutate(ratio = apop / prol)
```

```{r fig5b-table, echo = FALSE}
occupancy |>
  transmute(Schedule = label,
            `Max apoptotic fraction` = round(max_apop, 3),
            `Proliferating depletion` = round(prol_depl, 3),
            `Apoptotic:proliferating AUC ratio` = round(ratio, 2),
            `Fold change vs baseline` = round(fold, 3)) |>
  arrange(desc(`Apoptotic:proliferating AUC ratio`)) |>
  knitr::kable()
```

The paper makes four quantitative statements about these panels; each is
checked below.

```{r fig5-assert}
# "All groups achieved apoptotic fractions > 50% during the treatment period"
stopifnot(nrow(occupancy) == 7L, all(occupancy$max_apop > 0.50))

# "Both 37.5 mg/kg groups (continuous, 5/2) sustained an apoptotic fraction < 60%"
low_dose <- occupancy$max_apop[occupancy$grp %in% c(7L, 8L)]
stopifnot(length(low_dose) == 2L, all(low_dose < 0.60))

# "Sustained regression occurred in individual mice under most intermittent
#  75 mg/kg regimens, but not at a 3 day on/4 day off schedule or at the
#  37.5 mg/kg dose level" -- exactly the > 2:1 apoptotic:proliferating split.
sustained <- occupancy$ratio[occupancy$grp %in% c(2L, 3L, 4L, 5L)]
not_sustained <- occupancy$ratio[occupancy$grp %in% c(6L, 7L, 8L)]
stopifnot(length(sustained) == 4L, length(not_sustained) == 3L)
stopifnot(min(sustained) > 2, max(not_sustained) < 2)

# "Linear relationships were observed between efficacy [...] and fractional
#  populations in proliferative (R2 = 0.97) and apoptotic (R2 = 0.93)
#  fractions, but not for the quiescent population (R2 = 0.42)"
r2 <- c(proliferating = cor(occupancy$prol,   occupancy$fold)^2,
        apoptotic     = cor(occupancy$apop,   occupancy$fold)^2,
        quiescent     = cor(occupancy$quiesc, occupancy$fold)^2)
round(r2, 3)
stopifnot(r2[["quiescent"]] < r2[["apoptotic"]],
          r2[["quiescent"]] < r2[["proliferating"]],
          abs(r2[["quiescent"]] - 0.42) < 0.10)
```

```{r fig5b, echo = FALSE, fig.cap = "Replicates Figure 5b of DeJongh 2025: efficacy (fold change from baseline 24 h after the last dose) against the fractional-population AUC of each cell state."}
occupancy |>
  select(label, fold, Proliferating = prol, Quiescent = quiesc, Apoptotic = apop) |>
  pivot_longer(-c(label, fold), names_to = "state", values_to = "auc") |>
  ggplot(aes(auc, fold)) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              colour = "firebrick", linewidth = 0.5) +
  geom_point(size = 2) +
  facet_wrap(~state, scales = "free_x") +
  labs(x = "Fractional population AUC over the treatment period",
       y = "Fold change in tumour volume") +
  theme_bw()
```

The ordering the paper reports is recovered: proliferating and apoptotic
fractions both track efficacy closely, the quiescent fraction much less so, and
the model's quiescent R2 lands within 0.03 of the published 0.42. The
apoptotic:proliferating ratio separates the four schedules the paper reports as
achieving sustained regression from the three it reports as failing, at exactly
the published 2:1 threshold.

## Check 5 - the tumour static equilibrium (Figure S10)

The supplementary derivation gives the tumour-static-equilibrium condition in
closed form. Writing the two drug effects as `w = 1 + EColap * C_olap` and
`a = 1 - C_az / (C_az + EC50az)`, stasis at negligible baseline volume requires
`w = 1 + a`, i.e.

`EColap * C_olap = 1 - C_az / (C_az + EC50az)`.

```{r tse}
ECOLAP <- 0.0154; EC50AZ <- 8.99
tse_olaparib <- function(c_az) (1 - c_az / (c_az + EC50AZ)) / ECOLAP
published <- tibble::tribble(
  ~Claim,                                        ~Published, ~Model,
  "Olaparib alone, TSE concentration (uM)",           64.9,  tse_olaparib(0),
  "Olaparib TSE at 20.2 uM AZD7648 (uM)",             19.9,  tse_olaparib(20.2)
)
stopifnot(with(published, max(abs(Model - Published) / Published)) < 0.01)
```

```{r tse-table, echo = FALSE}
published |>
  mutate(across(c(Published, Model), \(x) round(x, 2)),
         `Difference (%)` = round(100 * (Model - Published) / Published, 1)) |>
  rename("Quantity" = Claim, "DeJongh 2025" = Published, "Packaged model" = Model) |>
  knitr::kable()
```

AZD7648 monotherapy can never reach the equilibrium: with `C_olap = 0` the
left-hand side is 0 while the right-hand side is strictly positive for any
finite AZD7648 concentration. That reproduces the paper's statement that
"AZD7648 is unable to achieve TSE as a monotherapy".

```{r fig-s10, echo = FALSE, fig.cap = "Replicates Figure S10 of DeJongh 2025: the tumour static equilibrium nomogram. Concentration pairs above the curve drive regression; pairs below it are insufficient for stasis."}
tibble::tibble(c_az = seq(0, 60, by = 0.5)) |>
  mutate(c_olap = tse_olaparib(c_az)) |>
  ggplot(aes(c_az, c_olap)) +
  geom_line(colour = "goldenrod3", linewidth = 1) +
  geom_point(data = tibble::tibble(c_az = c(0, 20.2), c_olap = c(64.9, 19.9)),
             colour = "black", size = 2.5) +
  labs(x = "AZD7648 steady-state concentration (uM)",
       y = "Olaparib steady-state concentration (uM)") +
  theme_bw()
```

# Assumptions and deviations

**Errata against the published equations.** Three printed equations disagree
with the NONMEM control streams the authors supplied. In each case the control
stream is authoritative and is what the packaged models implement, because the
control streams are what produced the published parameter estimates and
figures.

1. **Equation 1, Michaelis-Menten elimination.** As printed, the saturable term
   is `Vmax / (A1/Vc + Km)`, which has units of a clearance rather than a rate
   and would not balance the equation. Both AZD7648 control streams (supplementary
   files 4 and 5) code `Vmax*A(2)/(A(2)/Vc + Km)`. The amount in the numerator
   was dropped in typesetting.

2. **Equation 16, the olaparib effect.** As printed, `OlEff = Ce,Olap * EColap`.
   The control stream codes `OLEFF = 1 + ECol*COL`. The leading 1 is the
   untreated baseline flux between the proliferating and quiescent states;
   without it vehicle animals would have no cell cycling at all, the untreated
   steady state derived in supplementary file 2 would not exist, and the
   40/20/20/20 split reproduced in Check 4 above would be unreachable.

3. **Equation 10, the logistic limitation term.** The printed form reads as
   `1 - (1 - Kto/2) * TXeno/KC`. The control stream codes `1 - Atot/(2*KC)`,
   which is also the form used in the supplementary steady-state derivation.

**Residual numerical differences.** Two Figure 5a numbers do not match exactly.
The paper reports a maximum apoptotic fraction "> 70%" for the 75 mg/kg 7 on /
7 off schedule and 67% for the 3 on / 4 off schedule; the packaged model gives
66% and 57%. Every other statement about that panel reproduces (all groups
above 50%, both 37.5 mg/kg groups below 60%, proliferating depletion above 90%
for the continuous / 14-14-14 / 5-2 schedules). The 3 on / 4 off group is also
the one the paper singles out as its poorest forecast, so some of the gap is
likely the same phenomenon. No parameter was adjusted to close it.

**Table labelling.** Table 2's "Absolute bio-avail F1 = 0.598" row is the
study S1734 SCID factor (control stream `THETA(10)`); the value anchored on the
intravenous arm is the 0.425 entry labelled "F1 11,448 (Nude)"
(`THETA(7)`, commented "abs F1, based on 11448"). Table 3's `Kto` row carries
the abbreviation and units but no value, with 0.243 printed on the following
"Proportional residual error" line; footnote c supplies `Kto = GC`.

**Residual-error scale.** Tables 1 and 2 label the proportional residual error
as a variance and the control streams confirm it (`W = IPRED*SQRT(THETA)`), so
those models use `sqrt(0.304)` and `sqrt(0.399)`. Table 3 gives no such label,
and its control stream uses `W = IPRED*THETA(9)` with `$SIGMA 1 FIX`, so 0.243
is a standard deviation and is used unchanged. Getting this wrong in either
direction would misstate the residual magnitude by roughly a factor of two.

**Unit conventions carried as published.** Olaparib clearance and
intercompartmental clearance are reported in L/h while its volumes are in L/kg.
The rate constants the model actually integrates are unaffected, and the scale
mismatch is absorbed into the per-study bioavailability factors - which is why
some exceed 1 (S1770 = 1.48, S1816 = 4.18). These are descriptive inter-study
scaling factors, not physiological bioavailabilities. Similarly, the AZD7648
`Vmax` of 4.27 umol/kg/h enters the saturable term multiplied by the central
amount rather than the concentration, so it is not a maximum elimination rate
in the usual dimensional sense; it is transcribed exactly as coded.

**Assumptions the source made explicit.** `Vp = Vc` for AZD7648 (Table 1);
`ka = 9.9 1/h` for AZD7648 in nude mice and `ka = 10 1/h` for olaparib, both
fixed for want of absorption-phase data; `Kto = GC` (Table 3 footnote c); and
all tumour cells proliferating at study start, which the control stream notes
has "very limited sensitivity on GoF".

**Assumptions made here.** The molar conversions (AZD7648 380.4 g/mol,
olaparib 435.1 g/mol) are not stated in the paper text; they are derived
exactly from the paired `AMT` (umol/kg) and `DOSE` (mg/kg) columns of the
supplied NONMEM dataset. Figure 1 is reproduced with the study S1734 SCID
bioavailability factor, since that is the source study for single-dose SCID
olaparib data. The observed medians plotted against the model are computed from
the supplied dataset rather than digitised from the figures. The tumour PK-PD
model has no PK random effects by construction, so the PK feeding the tumour
system is the typical-value profile for every animal - exactly as the authors
fitted it.
