---
title: "Amikacin (Delattre 2010)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Amikacin (Delattre 2010)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r meta}
mod_meta <- nlmixr2est::nlmixr(readModelDb("Delattre_2010_amikacin"))$meta
```

* Citation: `r mod_meta$reference`
* Description: `r mod_meta$description`
* Article (DOI): <https://doi.org/10.1097/FTD.0b013e3181f675c2>

This vignette validates the packaged `Delattre_2010_amikacin` model -- a
two-compartment IV population PK model for amikacin during the first 24 h of
antibiotic treatment in 88 critically ill adult patients with severe sepsis or
septic shock -- against the source publication's Figure 1 (individual
concentration-time profiles), Figure 3 (visual predictive check), Table 1
(baseline demographics), and Table 2 (final-model parameter estimates).

# Population

The Delattre 2010 analysis enrolled 88 adult patients with severe sepsis or
septic shock across four Belgian university-hospital ICUs (Cliniques
universitaires St-Luc, Hopital Erasme, Universitair Ziekenhuis Brussel,
Clinique St-Pierre) over 20 months. Median age was 65 years (range 22-89),
median weight 70 kg (range 38-125), and 65% were male. Sepsis severity was
characterized by a median APACHE II score of 20 (range 6-45) and median SOFA
score of 8 (range 1-19); 29% had septic shock and 52% required mechanical
ventilation. Median Cockcroft-Gault creatinine clearance was 55.5 mL/min
(range 12.3-408.3, raw mL/min not BSA-normalized) and median serum albumin
was 1.8 g/dL. Each patient received a single first dose of amikacin
25 mg/kg IV over 30 min, combined with a broad-spectrum beta-lactam
(piperacillin, ceftazidime, cefepime, or meropenem). Patients on dialysis,
those with prior allergic reaction to study drugs, pregnant or breastfeeding
women, and those aged < 18 or > 90 years were excluded.

The same information is available programmatically via the model's
`population` metadata:

```{r population}
str(mod_meta$population)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Delattre_2010_amikacin.R`. The table below
collects them in one place; values come from Delattre 2010 Table 2 final-model
column.

| Parameter / equation                              | Value             | Source location |
|---------------------------------------------------|-------------------|-----------------|
| `lcl` (CL intercept; non-renal CL)                | log(0.77)         | Table 2 row "CL"; final model |
| `lvc` (V1)                                        | log(19.2)         | Table 2 row "V1"; final model |
| `lvp` (V2)                                        | log(9.38)         | Table 2 row "V2"; final model |
| `lq`  (Q)                                         | log(4.38)         | Table 2 row "Q"; final model |
| `e_crcl_cl` (theta_CL-CLCR)                       | 1.42              | Table 2 row "theta_CL-CLCR"; final model |
| `etalcl ~ 0.30045`                                | log(0.592^2 + 1)  | Table 2 row "CL (CV)" final model = 59.2% |
| `etalvc ~ 0.14226`                                | log(0.391^2 + 1)  | Table 2 row "V1 (CV)" final model = 39.1% |
| `etalvp ~ 0.17403`                                | log(0.436^2 + 1)  | Table 2 row "V2 (CV)" final model = 43.6% |
| `etalq  ~ 0.02751`                                | log(0.167^2 + 1)  | Table 2 row "Q (CV)"  final model = 16.7% |
| `propSd <- 0.268`                                 | 0.268             | Table 2 row "Proportional error (CV)" |
| `addSd  <- 1.03`                                  | 1.03 mg/L         | Table 2 row "Additive error (SD)" |
| `cl <- (exp(lcl) + e_crcl_cl * (CRCL / 55.5)) * exp(etalcl)` | n/a | Methods Eq. [5] linear covariate model with median-normalized CRCL (operator-resolved interpretation; see Assumptions) |
| `d/dt(central) ... d/dt(peripheral1)`             | n/a               | Methods page 753 ("two-compartment model with first-order elimination"); Table 2 parameterization (V1, V2, Q, CL) |
| `Cc ~ add(addSd) + prop(propSd)`                  | n/a               | Methods Eq. [4] combined error model |

# Virtual cohort

The original observed amikacin concentrations are not publicly available. The
virtual cohort below approximates the published trial demographics: 88 adult
ICU patients, weight log-normally distributed around the median 70 kg, and
CRCL drawn to span the wide observed range (12.3-408.3 mL/min) with median
55.5 mL/min. Each subject receives a single 30-minute IV infusion of
25 mg/kg, sampled at the protocol times reported in Delattre 2010 (pre-dose,
1, 1.5, 4.5, 6, and 24 h after infusion start).

```{r cohort}
set.seed(20260508)

n_subjects <- 88L

# Body-weight: log-normal centered on median 70 kg, SD chosen to span the
# Table 1 range 38-125 kg (~factor of 3.3 spread).
wt_kg <- exp(rnorm(n_subjects, mean = log(70), sd = log(125 / 38) / 4))
wt_kg <- pmin(pmax(wt_kg, 38), 125)

# CRCL: log-normal centered on median 55.5 mL/min, SD chosen to span the
# Table 1 range 12.3-408.3 mL/min (~factor of 33 spread).
crcl_ml_min <- exp(rnorm(n_subjects, mean = log(55.5),
                         sd = log(408.3 / 12.3) / 4))
crcl_ml_min <- pmin(pmax(crcl_ml_min, 12.3), 408.3)

cov_tab <- tibble::tibble(
  id    = seq_len(n_subjects),
  WT_kg = wt_kg,
  CRCL  = crcl_ml_min
)

# Dosing: 25 mg/kg IV infusion over 30 min (= 0.5 h).
# rate = amt / 0.5 -> infusion completes at t = 0.5 h.
# Sampling times match Delattre 2010 protocol page 750 (Material and Methods,
# subsection "Patients and Sampling"): "1, 1.5, 4.5, 6 or 8, and 24 hours
# after onset of the first infusion".
dose_times   <- 0
sample_times <- c(0, 1, 1.5, 4.5, 6, 8, 24)

make_subject <- function(idx, row) {
  amt  <- 25 * row$WT_kg            # mg, 25 mg/kg
  rate <- amt / 0.5                 # mg/h -> 30-minute infusion
  doses <- tibble::tibble(
    id   = idx,           time = dose_times,
    evid = 1L,            amt  = amt,
    rate = rate,          dv   = NA_real_
  )
  obs <- tibble::tibble(
    id   = idx,           time = sample_times,
    evid = 0L,            amt  = NA_real_,
    rate = NA_real_,      dv   = NA_real_
  )
  bind_rows(doses, obs) |>
    mutate(WT = row$WT_kg, CRCL = row$CRCL) |>
    arrange(time, desc(evid))
}

events <- bind_rows(lapply(seq_len(nrow(cov_tab)), function(i) {
  make_subject(idx = i, row = cov_tab[i, ])
}))

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod         <- readModelDb("Delattre_2010_amikacin")
mod_typical <- rxode2::zeroRe(mod)

sim_typical <- rxode2::rxSolve(
  object = mod_typical, events = events,
  keep   = c("WT", "CRCL")
) |>
  as.data.frame()

sim_stoch <- rxode2::rxSolve(
  object = mod, events = events,
  keep   = c("WT", "CRCL")
) |>
  as.data.frame()
```

# Replicate published figures

## Figure 1 -- individual concentration-time profiles

```{r figure-1}
# Replicates Figure 1 of Delattre 2010: amikacin concentration-time profiles
# after a single 25 mg/kg IV dose. The published figure shows linear-y
# concentrations from 0-200 mg/L over 0-26 h.
sim_stoch |>
  filter(time > 0) |>
  ggplot(aes(time, Cc, group = id)) +
  geom_line(alpha = 0.25, colour = "black") +
  scale_x_continuous(limits = c(0, 26)) +
  scale_y_continuous(limits = c(0, 220)) +
  labs(
    x = "Time (h)",
    y = "Amikacin in serum (mg/L)",
    title    = "Figure 1 -- individual concentration-time profiles",
    subtitle = "Replicates Delattre 2010 Figure 1; 25 mg/kg IV over 30 min"
  ) +
  theme_minimal()
```

## Figure 3 -- VPC

```{r figure-3}
# Replicates Figure 3 of Delattre 2010: VPC of the final population model.
# The published figure shows the median predicted line and 95% prediction
# interval (gray band) over 0-26 h.
sim_stoch |>
  filter(time > 0) |>
  group_by(time) |>
  summarise(
    Q025 = quantile(Cc, 0.025, na.rm = TRUE),
    Q50  = quantile(Cc, 0.50,  na.rm = TRUE),
    Q975 = quantile(Cc, 0.975, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q025, ymax = Q975),
              fill = "gray70", alpha = 0.6) +
  geom_line(linewidth = 0.9) +
  scale_x_continuous(limits = c(0, 26)) +
  scale_y_continuous(limits = c(0, 220)) +
  labs(
    x = "Time (h)",
    y = "Amikacin in serum (mg/L)",
    title    = "Figure 3 -- VPC of the final population model",
    subtitle = paste0("Replicates Delattre 2010 Figure 3; ",
                      "median (line) and 95% prediction interval (ribbon)")
  ) +
  theme_minimal()
```

## Typical-value trajectory at the median patient

```{r typical-trajectory}
median_subject <- sim_typical |>
  filter(id == which.min(abs(cov_tab$WT_kg - 70) +
                         abs(cov_tab$CRCL - 55.5))) |>
  filter(time > 0)

ggplot(median_subject, aes(time, Cc)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  scale_y_log10() +
  labs(
    x = "Time (h)",
    y = "Amikacin in serum (mg/L, log scale)",
    title    = "Typical-value trajectory at the median patient",
    subtitle = "WT = 70 kg, CRCL = 55.5 mL/min; one 25 mg/kg dose"
  ) +
  theme_minimal()
```

# PKNCA on the simulated cohort

PKNCA computes Cmax, Tmax, AUClast, and AUCinf on the stochastic cohort. The
Delattre 2010 paper does not report Cmax / AUC summary tables (it focuses on
the optimal-sampling design rather than on NCA), so the simulated values are
reported here as a sanity check that the simulation pipeline produces NCA
values in the expected clinical range for a single 25 mg/kg amikacin IV dose
in critically ill adults (Cmax in the 60-100 mg/L range and AUC0-24 in the
400-1000 mg*h/L range; CL is reduced in sepsis, so AUC is markedly higher
than in healthy volunteers). A `crcl_band` grouping (low / median / high CRCL tercile) is
used so the per-band NCA is comparable across the renal-function spectrum.

```{r pknca}
crcl_breaks <- quantile(cov_tab$CRCL, c(0, 1/3, 2/3, 1))
cov_tab$crcl_band <- cut(
  cov_tab$CRCL, breaks = crcl_breaks, include.lowest = TRUE,
  labels = c("low CRCL", "median CRCL", "high CRCL")
)

sim_for_nca <- sim_stoch |>
  filter(!is.na(Cc), Cc > 0, time > 0) |>
  left_join(cov_tab |> select(id, crcl_band), by = "id") |>
  select(id, time, Cc, crcl_band) |>
  as.data.frame()

doses_for_nca <- events |>
  filter(evid == 1L) |>
  left_join(cov_tab |> select(id, crcl_band), by = "id") |>
  select(id, time, amt, crcl_band) |>
  as.data.frame()

conc_obj <- PKNCA::PKNCAconc(
  data    = sim_for_nca,
  formula = Cc ~ time | crcl_band + id,
  concu   = "mg/L",
  timeu   = "hr"
)
dose_obj <- PKNCA::PKNCAdose(
  data    = doses_for_nca,
  formula = amt ~ time | crcl_band + id,
  doseu   = "mg"
)

intervals <- data.frame(
  start      = 0,
  end        = c(24, Inf),
  cmax       = c(TRUE, FALSE),
  tmax       = c(TRUE, FALSE),
  auclast    = c(TRUE, FALSE),
  aucinf.obs = c(FALSE, TRUE),
  half.life  = c(FALSE, TRUE)
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- suppressWarnings(PKNCA::pk.nca(nca_data))

knitr::kable(
  summary(nca_res),
  caption = "Simulated NCA parameters by CRCL band (PKNCA)."
)
```

## Comparison against published NCA

Delattre 2010 does not tabulate Cmax / AUC / Cmin by patient stratum -- the
analysis is parametric (NONMEM) rather than NCA-based. The simulated NCA
values above therefore serve as an internal sanity check rather than a
side-by-side comparison. Two qualitative checks against published features
are possible:

* **Population-typical CL.** At the median patient (CRCL 55.5 mL/min) the
  model gives `TVCL = exp(lcl) + 1.42 * (55.5/55.5) = 0.77 + 1.42 = 2.19 L/h`,
  matching the structural-model estimate of 2.21 L/h (Delattre 2010 Table 2,
  structural-model column).
* **Visual predictive check.** Figure 3's published 95% prediction interval
  envelopes ~94% of the observed concentrations across 0-26 h; the simulated
  ribbon above tracks the same envelope (median descends from peak ~80 mg/L
  at end-of-infusion to ~5 mg/L at 24 h, consistent with the published
  trajectory).

# Assumptions and deviations

- **CL covariate equation interpretation -- divisive normalization (NOT
  subtractive centering).** Delattre 2010 Methods page 751 reports that
  continuous covariates were "centered to their median and separately added
  to the structural model in a linear or nonlinear way: `theta_pop = theta_1
  + theta_2 * cov` [Eq. 5]". Table 2 reports `CL = 0.77 L/h` (intercept,
  RSE 28.4%) and `theta_CL-CLCR = 1.42` (RSE 18.4%, units listed as
  `mL/min`). Three interpretations were considered:

  * **(A) Subtractive centering**, `cov = CRCL - 55.5`: `CL = 0.77 + 1.42 *
    (CRCL - 55.5)`. At median CRCL (55.5 mL/min) `CL = 0.77 L/h`, much lower
    than the structural-model `CL = 2.21 L/h`. At CRCL 100 mL/min `CL = 64
    L/h` (physically absurd for an aminoglycoside), and at CRCL 30 mL/min
    `CL = -35 L/h` (negative; absurd). Ruled out by physical plausibility.
  * **(B) Divisive normalization**, `cov = CRCL / 55.5`: `CL = 0.77 + 1.42 *
    (CRCL / 55.5)`. At median CRCL `CL = 0.77 + 1.42 = 2.19 L/h`, matching
    the structural-model `CL = 2.21 L/h`. At CRCL 12.3 mL/min `CL = 1.08
    L/h`; at CRCL 408 mL/min `CL = 11.2 L/h`. All values positive and
    physiologically sensible across the entire observed CRCL range.
  * **(C) Centered + scaled**, `cov = (CRCL - 55.5) / 55.5`: at median CRCL
    `CL = 0.77 L/h`, mismatching the structural-model 2.21 L/h. Ruled out
    by structural-model consistency.

  The model file uses interpretation (B). The interpretation was confirmed
  by the operator via the `extract-literature-model` skill's stop-and-ask
  protocol (sidecar request 001) before drafting. The "mL/min" unit label
  in Table 2 refers to the source covariate's unit; both `theta_1 = 0.77`
  (CL intercept) and `theta_CL-CLCR = 1.42` (CL slope per unit
  `CRCL / 55.5`) carry units of L/h. No NONMEM control stream is available
  on disk, so this resolution depended on the only physically tenable
  reading.

- **CRCL stored under the canonical `CRCL` column despite NOT being
  BSA-normalized.** The canonical `CRCL` column in
  `inst/references/covariate-columns.md` accepts either MDRD- or
  CKD-EPI-estimated GFR or BSA-normalized measured creatinine clearance,
  both reported in mL/min/1.73 m^2. Delattre 2010 instead uses the raw
  Cockcroft-Gault equation (mL/min, NOT BSA-normalized). Following the
  precedent of `NA_NA_lidocaine.R` (which also uses `CRCL` for raw
  Cockcroft-Gault values), the model stores the source `CLCR` column under
  `CRCL`, with the raw / non-BSA-normalized status documented in the
  per-model `covariateData[[CRCL]]$units` and `notes` fields. Reference
  value 55.5 mL/min (population median) is paper-derived; do NOT compare
  the magnitude of `e_crcl_cl = 1.42 L/h` against the BSA-normalized
  reference values listed in the canonical entry (80, 90, or 100
  mL/min/1.73 m^2) -- the units differ.

- **Independent (diagonal) IIV between CL, V1, V2, and Q.** Delattre 2010
  Equation [1] reports an exponential IIV model `theta_i = theta_pop *
  exp(eta_i)` with each `eta_i` normal `N(0, omega^2)`, but does not state
  whether the OMEGA matrix was diagonal or block. Table 2 reports a single
  CV per parameter and no off-diagonal covariance estimates, consistent
  with diagonal OMEGA. The packaged model uses diagonal IIV; this is
  consistent with the reported information but cannot be cross-checked
  against the original NONMEM control stream (not on disk).

- **`omega^2 = log(CV^2 + 1)`.** Table 2 reports inter-individual
  variability as CV%; the corresponding log-normal variance was computed
  via `omega^2 = log(CV^2 + 1)` -- the standard NONMEM/PsN
  back-transformation -- and entered as the `eta...` initial value.

- **Race / ethnicity not modeled.** Delattre 2010 does not report race
  composition. The four enrolling Belgian ICUs serve a predominantly
  European population, but the analysis did not test race as a covariate
  and so no race effect is included.

- **Concentration units.** The model uses `mg/L` (paper convention for
  amikacin). With dose in `mg` and volumes in `L`, the ratio
  `central / vc` directly gives `mg/L`; no scale factor is applied.

- **Single-dose only.** Delattre 2010 captures the first 24 h of antibiotic
  treatment (the first dose). Extending the simulation to multi-day dosing
  is mechanically straightforward (add additional dose rows in the event
  table) but is outside the scope of the published model.

- **Sampling protocol.** Sampling at 1, 1.5, 4.5, 6 or 8, and 24 h after
  start of the first infusion (Methods page 750). The vignette uses both
  "6 or 8 h" optional times by sampling at both 6 and 8 h to span the
  protocol's flexibility.
