---
title: "Magnesium sulfate (Deng 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Magnesium sulfate (Deng 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Deng_2024_magnesiumSulfate"))$reference`
* Description: `r rxode2::rxode(readModelDb("Deng_2024_magnesiumSulfate"))$description`
* Article: <https://doi.org/10.1186/s12884-024-06620-x>

Deng 2024 is a prospective observational population-PK study of intravenous
magnesium sulfate for seizure prophylaxis in Chinese women with preeclampsia.
It is the third magnesium sulfate popPK model in this library, alongside
`Salinger_2013_magnesiumSulfate` (Indian cohort, intramuscular and intravenous
maintenance) and `Easterling_2018_magnesium_sulfate` (Egyptian cohort,
continuous infusion versus serial bolus). Its distinguishing contribution is
the covariate model: creatinine clearance, body mass index and concomitant
**furosemide** on clearance, plus furosemide on volume. The paper presents the
furosemide effect as the first such report for magnesium.

**Read the Assumptions and deviations section before using this model.** Two
features of the source publication required documented interpretation: the
normalising constants printed in the paper's covariate equation are not
usable as printed, and the published model carries no endogenous magnesium
baseline even though the data it was fit to are total serum magnesium.

# Population

```{r population-metadata}
pop <- rxode2::rxode(readModelDb("Deng_2024_magnesiumSulfate"))$population
str(pop, max.level = 1)
```

51 women contributed 199 serum magnesium concentrations (two to four samples
each) between April 2021 and April 2023 at the Affiliated Suzhou Hospital of
Nanjing Medical University. Baseline characteristics are Deng 2024 Table 1:

| Characteristic | Value |
|---|---|
| Age (years) | 31 (28, 35); mean 31.86 +/- 5.22 |
| Gestational age (weeks) | 32.31 +/- 3.92 |
| BMI (kg/m^2) | 29.13 (27.16, 33.30) |
| Baseline magnesium (mmol/L) | 0.76 (0.71, 0.86) |
| Creatinine clearance (mL/min) | 182.18 +/- 67.15 |
| Albumin (g/L) | 30.45 +/- 4.66 |
| Calcium (tabulated as g/L) | 2.17 +/- 0.15 |
| Twin pregnancy | 4 (7.84%) |
| Labetalol | 50 (98.03%) |
| Nifedipine | 22 (43.14%) |
| Furosemide | 16 (31.37%) |
| Any adverse reaction | 21 (41.18%) |

The cohort is renally hyperfiltrating: a mean creatinine clearance of
182 mL/min reflects the 40-65% pregnancy-associated rise in glomerular
filtration rate the paper describes. Any user applying this model outside a
comparable pregnant population is extrapolating the `(CRCL/175)^0.39` term
well below the range in which it was estimated.

Dosing in the study was 5 g MgSO4-7H2O intravenously over 30-120 min on day 1
followed by a 10 g maintenance infusion over 6-8 h, then the 10 g maintenance
dose alone on days 2-5. **All modelled samples were drawn on day 2**, after a
maintenance dose with no same-day loading dose.

# Unit conventions

Two conversions are load-bearing and are worth stating explicitly because the
paper leaves both implicit.

**Salt basis.** Doses are prescribed in grams of magnesium sulfate. Following
the two sibling models in this library, the salt is taken to be the
heptahydrate MgSO4-7H2O (MW 246.47) used in obstetric practice, and doses are
expressed in this model as mg of *elemental* magnesium (MW 24.305).

**Concentration.** Deng 2024 Figure 3 labels its visual-predictive-check
y-axis "Prediction-corrected concentrations (mg/L)", with observations
spanning roughly 20-52 mg/L. That fixes the concentration unit as mg/L of
elemental magnesium, and therefore fixes the additive residual SD of 3.65 as
3.65 mg/L (about 7-18% of the observed concentration range, a reasonable
magnitude for a clinical-chemistry assay). Table 1 and Figure 4 report
magnesium in mmol/L; 1 mmol/L = 24.305 mg/L.

```{r units}
MG_MW      <- 24.305                    # elemental Mg, g/mol
MGSO4_MW   <- 246.47                    # MgSO4-7H2O, g/mol
MG_PER_G   <- MG_MW / MGSO4_MW * 1000   # mg elemental Mg per g MgSO4-7H2O
BASELINE   <- 0.76 * MG_MW              # reported baseline magnesium, mg/L

c(mg_Mg_per_g_salt = MG_PER_G,
  g5  = 5  * MG_PER_G,
  g10 = 10 * MG_PER_G,
  g15 = 15 * MG_PER_G,
  baseline_mgL = BASELINE)
```

# Source trace

Every `ini()` value and every `model()` equation, with its origin in the
source publication. In-file comments next to each `ini()` entry carry the same
information.

| Model quantity | Value | Source location |
|---|---|---|
| `lcl` (CL) | 2.98 L/h | Table 2, `tvCL`; 95% CI 1.29-4.62. Also Abstract and Discussion. |
| `lvc` (V) | 25.07 L | Table 2, `tvV`; 95% CI 23.31-26.82 |
| `e_crcl_cl` | +0.39 | Table 2, `dCLdCCR`; 95% CI 0.31-0.46 |
| `e_bmi_cl` | -0.54 | Table 2, `dCLdBMI`; 95% CI (-0.70)-(-0.38) |
| `e_conmed_furosemide_cl` | -0.16 | Table 2, `dCLdfurosemide`; 95% CI (-0.22)-(-0.096) |
| `e_conmed_furosemide_vc` | -0.25 | Table 2, `dVdfurosemide`; 95% CI (-0.34)-(-0.16) |
| `etalcl` | 0.082 (variance) | Table 2, `omega^2 CL`; shrinkage 13.05% |
| `etalvc` | 0.023 (variance) | Table 2, `omega^2 V`; shrinkage 13.04% |
| `addSd` | 3.65 mg/L | Table 2, `stdev0`; 95% CI 3.20-4.09 |
| CL equation form | power CRCL, power BMI, fractional furosemide, `exp(eta)` | Results, "The equations for V and CL were as follows" |
| V equation form | fractional furosemide, `exp(eta)` | Results, same equation block |
| One-compartment disposition | -- | Results, "adequately described by one-compartment structural model" |
| Additive residual error | -- | Results, "an additive error model was used to evaluate the residual variability" |
| CRCL reference 175 mL/min | **deviation** | Monte Carlo section, "CCR and BMI are taken at the median (CCR:175 ml/min, BMI:29 kg/m2)". NOT the printed constant -- see Assumptions and deviations. |
| BMI reference 29 kg/m^2 | **deviation** | Monte Carlo section, same sentence. NOT the printed constant. |
| Concentration unit mg/L | -- | Figure 3 y-axis |
| Baseline magnesium 0.76 mmol/L | not a model parameter | Table 1 and Results; see Assumptions and deviations |

# Virtual cohort

Two arms of 100 women each, matching the covariate distributions in Table 1,
split by furosemide use. Creatinine clearance is drawn log-normally with the
Table 1 mean and SD; BMI is drawn log-normally from the Table 1 median and
interquartile range.

```{r cohort}
rxode2::rxSetSeed(20240424)
set.seed(20240424)

n_arm <- 100L

# CRCL: mean 182.18, SD 67.15 (Table 1) matched on the log scale
crcl_cv     <- 67.15 / 182.18
crcl_sdlog  <- sqrt(log(1 + crcl_cv^2))
crcl_meanlog <- log(182.18) - crcl_sdlog^2 / 2

# BMI: median 29.13, IQR 27.16-33.30 (Table 1)
bmi_sdlog   <- log(33.30 / 27.16) / (2 * qnorm(0.75))
bmi_meanlog <- log(29.13)

make_arm <- function(furo, offset) {
  data.frame(
    id                = offset + seq_len(n_arm),
    CRCL              = rlnorm(n_arm, crcl_meanlog, crcl_sdlog),
    BMI               = rlnorm(n_arm, bmi_meanlog, bmi_sdlog),
    CONMED_FUROSEMIDE = furo,
    arm               = if (furo == 1) "Furosemide" else "No furosemide"
  )
}

cohort <- dplyr::bind_rows(make_arm(0, 0L), make_arm(1, n_arm))

cohort |>
  dplyr::group_by(arm) |>
  dplyr::summarise(n = dplyr::n(),
                   CRCL_median = median(CRCL), CRCL_mean = mean(CRCL),
                   CRCL_sd = sd(CRCL), BMI_median = median(BMI),
                   .groups = "drop") |>
  knitr::kable(digits = 2,
               caption = "Simulated covariate distributions against Table 1 targets (CRCL mean 182.18 +/- 67.15 mL/min; BMI median 29.13 kg/m^2).")
```

The regimen simulated below is the one Deng 2024 recommends for women not
receiving furosemide and reports in Figure 4 and Figure 5 panel (d): a 5 g
loading dose over 20 minutes followed by a 10 g maintenance dose over 6.7 h.
Observations run to 24 h so that the terminal phase is well characterised for
non-compartmental analysis.

```{r events}
T_LOAD  <- 20 / 60      # 20 min loading infusion
T_MAINT <- 6.7          # 6.7 h maintenance infusion
AMT_LOAD  <- 5  * MG_PER_G
AMT_MAINT <- 10 * MG_PER_G
T_END <- T_LOAD + T_MAINT

times_obs <- sort(unique(c(
  seq(0, 1, by = 0.05),
  seq(1.25, 8, by = 0.25),
  T_LOAD, T_END,
  seq(8.5, 24, by = 0.5)
)))

obs_rows <- cohort |>
  tidyr::expand_grid(time = times_obs) |>
  dplyr::mutate(evid = 0L, amt = 0, dur = 0, cmt = "central")

dose_rows <- dplyr::bind_rows(
  cohort |> dplyr::mutate(time = 0,      evid = 1L, amt = AMT_LOAD,  dur = T_LOAD,  cmt = "central"),
  cohort |> dplyr::mutate(time = T_LOAD, evid = 1L, amt = AMT_MAINT, dur = T_MAINT, cmt = "central")
)

events <- dplyr::bind_rows(obs_rows, dose_rows) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(
  !anyDuplicated(events[events$evid == 1L, c("id", "time")]),
  all(c("CRCL", "BMI", "CONMED_FUROSEMIDE") %in% names(events))
)
```

# Simulation

`rxSolve()` on an `rxUi` object scales super-linearly in the number of
subjects, so each arm is solved in its own call.

```{r simulate}
mod <- readModelDb("Deng_2024_magnesiumSulfate")

solve_arm <- function(model, arm_name) {
  ev <- events[events$arm == arm_name, ]
  rxode2::rxSolve(model, events = ev,
                  keep = c("CRCL", "BMI", "CONMED_FUROSEMIDE", "arm"),
                  returnType = "data.frame")
}

# Typical-value predictions: zeroRe() removes both IIV and residual error.
mod_typ <- rxode2::zeroRe(mod)
sim_typ <- dplyr::bind_rows(
  solve_arm(mod_typ, "No furosemide"),
  solve_arm(mod_typ, "Furosemide")
)

# Stochastic cohort: IIV retained, residual error suppressed so that Cc is the
# individual prediction.
rxode2::rxSetSeed(20240424)
mod_iiv <- rxode2::zeroRe(mod, which = "sigma")
sim_iiv <- dplyr::bind_rows(
  solve_arm(mod_iiv, "No furosemide"),
  solve_arm(mod_iiv, "Furosemide")
)

nrow(sim_typ)
```

# Verification gate 1: closed-form solution

The model is a one-compartment system receiving two sequential zero-order
infusions, so it has a closed-form solution. Comparing the `rxode2` solve
against that solution for the reference subject (CRCL 175 mL/min, BMI
29 kg/m^2, no furosemide) tests the implementation with no confounding from
covariates or random effects. Both sides use the same parameter values, so the
only difference is integrator error and the tolerance is set accordingly.

```{r gate-closed-form}
CL_REF <- 2.98
V_REF  <- 25.07
K_REF  <- CL_REF / V_REF

conc_two_infusions <- function(t, cl, v, amt_load, t_load, amt_maint, t_maint) {
  k  <- cl / v
  rL <- amt_load / t_load
  rM <- amt_maint / t_maint
  c_load_end  <- (rL / cl) * (1 - exp(-k * t_load))
  c_maint_end <- (rM / cl) * (1 - exp(-k * t_maint)) + c_load_end * exp(-k * t_maint)
  ifelse(
    t <= t_load,
    (rL / cl) * (1 - exp(-k * t)),
    ifelse(
      t <= t_load + t_maint,
      (rM / cl) * (1 - exp(-k * (t - t_load))) + c_load_end * exp(-k * (t - t_load)),
      c_maint_end * exp(-k * (t - t_load - t_maint))
    )
  )
}

ref_ev <- rxode2::et(amt = AMT_LOAD,  dur = T_LOAD,  cmt = "central", time = 0) |>
  rxode2::et(amt = AMT_MAINT, dur = T_MAINT, cmt = "central", time = T_LOAD) |>
  rxode2::et(times_obs, cmt = "central") |>
  as.data.frame()
ref_ev$id <- 1L
ref_ev$CRCL <- 175
ref_ev$BMI <- 29
ref_ev$CONMED_FUROSEMIDE <- 0

ref_sim <- rxode2::rxSolve(mod_typ, ref_ev, returnType = "data.frame")
ref_obs <- ref_sim[!is.na(ref_sim$Cc), ]
ref_obs$analytic <- conc_two_infusions(
  ref_obs$time, CL_REF, V_REF, AMT_LOAD, T_LOAD, AMT_MAINT, T_MAINT
)

closed_form_err <- max(abs(ref_obs$Cc - ref_obs$analytic) /
                         pmax(ref_obs$analytic, 1e-8))
closed_form_err

# Same parameters on both sides: this is pure numerical error, so assert tightly.
stopifnot(closed_form_err < 1e-6)
```

The reference subject's clearance and volume must also come back exactly as
Table 2 reports them, confirming that the covariate term is 1 at the reference
covariate values.

```{r gate-reference-params}
ref_params <- unique(ref_sim[, c("cl", "vc", "kel")])
ref_params

stopifnot(
  abs(ref_params$cl  - CL_REF) < 1e-9,
  abs(ref_params$vc  - V_REF)  < 1e-9,
  abs(ref_params$kel - K_REF)  < 1e-9
)
```

# Verification gate 2: covariate model

Each covariate term is checked against the printed equation independently.
`(CRCL/175)^0.39`, `(BMI/29)^-0.54` and the two furosemide fractional-change
terms are all deterministic functions of the covariate vector, so exact
agreement is required.

```{r gate-covariates}
expected_cl <- function(crcl, bmi, furo) {
  CL_REF * (crcl / 175)^0.39 * (bmi / 29)^(-0.54) * (1 - 0.16 * furo)
}
expected_vc <- function(furo) V_REF * (1 - 0.25 * furo)

cov_check <- sim_typ |>
  dplyr::distinct(id, arm, CRCL, BMI, CONMED_FUROSEMIDE, cl, vc) |>
  dplyr::mutate(
    cl_expected = expected_cl(CRCL, BMI, CONMED_FUROSEMIDE),
    vc_expected = expected_vc(CONMED_FUROSEMIDE),
    cl_err = abs(cl - cl_expected) / cl_expected,
    vc_err = abs(vc - vc_expected) / vc_expected
  )

c(max_cl_err = max(cov_check$cl_err), max_vc_err = max(cov_check$vc_err))

stopifnot(max(cov_check$cl_err) < 1e-10, max(cov_check$vc_err) < 1e-10)
```

The furosemide effect must reproduce the Table 2 multipliers of 0.84 on
clearance and 0.75 on volume exactly.

```{r gate-furosemide}
furo_ratio <- cov_check |>
  dplyr::group_by(CONMED_FUROSEMIDE) |>
  dplyr::summarise(vc = unique(round(vc, 8)), .groups = "drop")

vc_ratio <- furo_ratio$vc[furo_ratio$CONMED_FUROSEMIDE == 1] /
  furo_ratio$vc[furo_ratio$CONMED_FUROSEMIDE == 0]

# Clearance ratio at matched covariates, isolating the furosemide term.
cl_ratio <- expected_cl(175, 29, 1) / expected_cl(175, 29, 0)

c(vc_ratio = vc_ratio, cl_ratio = cl_ratio)

stopifnot(abs(vc_ratio - 0.75) < 1e-8, abs(cl_ratio - 0.84) < 1e-8)
```

Both effects lower the parameter and therefore raise serum magnesium, matching
the paper's statement that "patients received furosemide had higher blood
magnesium concentrations than those who did not receive furosemide".

# Replicate published figures

## Figure 4: median concentration-time profiles

Figure 4 of Deng 2024 plots the median simulated magnesium concentration
against time for eight dosing regimens at median covariates, split by
furosemide use. Below we reproduce the two 10 g maintenance regimens for both
arms, on the paper's mmol/L axis and with the reported endogenous baseline
added so the curves are on the same footing as the published ones.

```{r figure-4, fig.width = 7, fig.height = 4.5}
fig4 <- sim_typ |>
  dplyr::filter(!is.na(Cc), time <= 8) |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(mmol = (median(Cc) + BASELINE) / MG_MW, .groups = "drop")

ggplot(fig4, aes(time, mmol, colour = arm)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = c(2, 3.5), linetype = "dashed", colour = "grey40") +
  scale_x_continuous(breaks = 0:8) +
  labs(x = "Time (hours)", y = "Magnesium concentration (mmol/L)",
       colour = NULL,
       title = "Figure 4 (replicated): 5 g load + 10 g maintenance",
       caption = paste("Dashed lines: the 2.0-3.5 mmol/L therapeutic window.",
                       "Endogenous baseline of 0.76 mmol/L added; see Assumptions.")) +
  theme(legend.position = "bottom")
```

The published curves for these two regimens read as follows at integer hours
(Figure 4, grey series). We compare them against the replication.

```{r figure-4-table}
published_fig4 <- tibble::tibble(
  time = 0:8,
  `No furosemide` = c(0.80, 2.08, 2.30, 2.49, 2.65, 2.77, 2.89, 2.99, 2.53),
  `Furosemide`    = c(0.80, 2.64, 2.89, 3.09, 3.27, 3.42, 3.53, 3.62, 3.01)
) |>
  tidyr::pivot_longer(-time, names_to = "arm", values_to = "published")

fig4_cmp <- sim_typ |>
  dplyr::filter(!is.na(Cc), time %in% 0:8) |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(modelled = (median(Cc) + BASELINE) / MG_MW, .groups = "drop") |>
  dplyr::inner_join(published_fig4, by = c("arm", "time")) |>
  dplyr::mutate(ratio_drug = (modelled - 0.76) / (published - 0.76))

fig4_cmp |>
  dplyr::filter(time > 0) |>
  dplyr::select(arm, time, modelled, published, ratio_drug) |>
  dplyr::rename("Arm" = arm, "Time (h)" = time,
                "Modelled (mmol/L)" = modelled,
                "Published (mmol/L)" = published,
                "Modelled/published, drug only" = ratio_drug) |>
  knitr::kable(digits = 3,
               caption = "Replication of Deng 2024 Figure 4 at integer hours. The last column divides the baseline-subtracted concentrations.")
```

The modelled curve sits systematically **below** the published one by a
near-constant factor on the drug-attributable concentration. That factor is
reported rather than corrected; see Assumptions and deviations, which shows
that the discrepancy is in amplitude only and not in shape.

## Figure 4 shape: the diagnostic that fixes CL/V

Because the amplitude of Figure 4 is not reproducible, the useful information
in it is its *shape*. If the modelled and published profiles differ only by a
constant scale factor, then the ratio `modelled / published` of the
baseline-subtracted concentrations is flat across time, and the elimination
rate constant `k = CL/V` is right even though the amplitude is not. If `k` is
wrong, that ratio drifts systematically.

Using all seven hourly points makes this robust to the roughly +/- 0.02 mmol/L
uncertainty of reading values off a raster figure. (A curvature statistic
built from differences between adjacent points -- for example
`(C7-C5)/(C5-C3)` -- is *not* robust: the differences involved are only
0.2-0.3 mmol/L, so a 0.02 mmol/L reading error moves the statistic by more
than 0.1 and the two arms give conflicting answers. It is not used here.)

```{r figure-4-shape}
cl_as_printed <- 2.98 * (175 / 51)^0.39 * (29 / 51)^(-0.54)

shape_ratio <- function(cl, vc, published_drug) {
  modelled <- conc_two_infusions(1:7, cl, vc,
                                 AMT_LOAD, T_LOAD, AMT_MAINT, T_MAINT) / MG_MW
  modelled / published_drug
}

shape_one_arm <- function(a) {
  furo <- as.integer(a == "Furosemide")
  pub  <- published_fig4$published[published_fig4$arm == a &
                                     published_fig4$time %in% 1:7] - 0.76
  vc   <- V_REF * (1 - 0.25 * furo)
  impl <- shape_ratio(2.98          * (1 - 0.16 * furo), vc, pub)
  litl <- shape_ratio(cl_as_printed * (1 - 0.16 * furo), vc, pub)
  data.frame(
    arm = a,
    implemented_mean = mean(impl), implemented_cv = 100 * sd(impl) / mean(impl),
    as_printed_mean  = mean(litl), as_printed_cv  = 100 * sd(litl) / mean(litl)
  )
}

shape_tbl <- dplyr::bind_rows(lapply(c("No furosemide", "Furosemide"),
                                     shape_one_arm))

shape_tbl |>
  dplyr::rename("Arm" = arm,
                "Ratio, implemented" = implemented_mean,
                "CV%, implemented"   = implemented_cv,
                "Ratio, as printed"  = as_printed_mean,
                "CV%, as printed"    = as_printed_cv) |>
  knitr::kable(digits = c(0, 3, 1, 3, 1),
               caption = "Modelled/published ratio of drug-attributable concentration at t = 1-7 h. A flat ratio (low CV%) means the shape matches and only the amplitude differs.")

c(cl_if_equation_read_literally = cl_as_printed)

# The implemented reading differs from Figure 4 by a near-constant factor in
# both arms; the literal reading of the printed equation drifts monotonically.
stopifnot(
  all(shape_tbl$implemented_cv < 4),
  all(shape_tbl$as_printed_cv > 3 * shape_tbl$implemented_cv)
)
```

Under the implemented reading the ratio is flat to within 1-2% in both arms,
so Figure 4 and this model agree on shape and disagree only on scale. Under
the literal reading of the printed equation the ratio falls monotonically
from about 0.61 to 0.40 across the same seven hours, a roughly ten-fold worse
agreement. This is the quantitative basis for the covariate-reference
decision documented below.

## Stochastic cohort profile

```{r cohort-profile, fig.width = 7, fig.height = 4.5}
band <- sim_iiv |>
  dplyr::filter(!is.na(Cc), time <= 16) |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(
    med = (median(Cc) + BASELINE) / MG_MW,
    lo  = (quantile(Cc, 0.05) + BASELINE) / MG_MW,
    hi  = (quantile(Cc, 0.95) + BASELINE) / MG_MW,
    .groups = "drop"
  )

ggplot(band, aes(time, med)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = c(2, 3.5), linetype = "dashed", colour = "grey40") +
  facet_wrap(~ arm) +
  labs(x = "Time (hours)", y = "Magnesium concentration (mmol/L)",
       title = "Simulated cohort, 5 g load + 10 g maintenance",
       caption = "Line = median; ribbon = 5th-95th centile across 100 women per arm.")
```

# PKNCA validation

Non-compartmental analysis of the typical-value profiles. Because the model
predicts administered magnesium only, no baseline correction is needed before
NCA. For a linear one-compartment model the NCA-derived clearance
(`Dose / AUC_inf`) and half-life must recover the model's own `CL` and
`ln(2) * V / CL` exactly, which makes this a strict check of the packaged
parameters rather than a descriptive summary.

```{r pknca}
conc_nca <- sim_typ |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

# Time-zero records are present by construction (times_obs starts at 0), which
# keeps PKNCA from warning about an AUC interval starting before the first
# measurement.
stopifnot(all(tapply(conc_nca$time, conc_nca$id, min) == 0))

dose_nca <- events |>
  dplyr::filter(evid == 1L) |>
  dplyr::group_by(id, arm) |>
  dplyr::summarise(time = min(time), amt = sum(amt), .groups = "drop")

conc_obj <- PKNCA::PKNCAconc(conc_nca, Cc ~ time | arm + id,
                             concu = "mg/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_nca, amt ~ time | arm + id, doseu = "mg")

intervals <- data.frame(
  start       = 0,
  end         = Inf,
  cmax        = TRUE,
  tmax        = TRUE,
  auclast     = TRUE,
  aucinf.obs  = TRUE,
  half.life   = TRUE,
  cl.obs      = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(arm, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

nca_wide |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    Cmax      = median(cmax),
    Tmax      = median(tmax),
    AUClast   = median(auclast),
    AUCinf    = median(aucinf.obs),
    half_life = median(half.life),
    CL_nca    = median(cl.obs),
    .groups   = "drop"
  ) |>
  dplyr::rename("Arm" = arm, "Cmax (mg/L)" = Cmax, "Tmax (h)" = Tmax,
                "AUClast (mg*h/L)" = AUClast, "AUCinf (mg*h/L)" = AUCinf,
                "t1/2 (h)" = half_life, "CL from NCA (L/h)" = CL_nca) |>
  knitr::kable(digits = 3,
               caption = "PKNCA results on the typical-value profiles, by arm.")
```

## Comparison against the model's own parameters

```{r pknca-gate}
nca_gate <- nca_wide |>
  dplyr::inner_join(
    sim_typ |> dplyr::distinct(id, cl, vc),
    by = "id"
  ) |>
  dplyr::mutate(
    cl_expected = cl,
    hl_expected = log(2) * vc / cl,
    cl_pct_diff = 100 * (cl.obs - cl_expected) / cl_expected,
    hl_pct_diff = 100 * (half.life - hl_expected) / hl_expected
  )

nca_gate |>
  dplyr::summarise(
    max_abs_cl_pct_diff = max(abs(cl_pct_diff)),
    max_abs_hl_pct_diff = max(abs(hl_pct_diff))
  )

# Both sides are built from the same parameter values, so the residual is
# trapezoidal / log-linear-regression error only and a tight bound is correct.
stopifnot(
  max(abs(nca_gate$cl_pct_diff)) < 1,
  max(abs(nca_gate$hl_pct_diff)) < 1
)
```

Deng 2024 reports no non-compartmental parameters, so there is no published
NCA table to compare against; the check above therefore verifies the packaged
model against its own closed-form exposure metrics. The reference subject's
values are:

```{r reference-exposure}
c(AUCinf_expected_mgh_per_L = (AMT_LOAD + AMT_MAINT) / CL_REF,
  half_life_expected_h      = log(2) * V_REF / CL_REF)
```

## Therapeutic target attainment

The paper's clinical endpoint is the fraction of women inside the
2.0-3.5 mmol/L therapeutic window. Its own observed figure is that 5 of 51
women (9.80%) reached the window at 4-5 h on day 2, when only the 10 g
maintenance dose was given and no loading dose preceded it. Simulating that
day-2 regimen gives the model's counterpart.

```{r target-attainment}
maint_only <- cohort |>
  tidyr::expand_grid(time = c(0, 4, 4.5, 5, 7)) |>
  dplyr::mutate(evid = 0L, amt = 0, dur = 0, cmt = "central") |>
  dplyr::bind_rows(
    cohort |> dplyr::mutate(time = 0, evid = 1L, amt = AMT_MAINT,
                            dur = 7, cmt = "central")
  ) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

rxode2::rxSetSeed(20240424)
sim_maint <- dplyr::bind_rows(
  rxode2::rxSolve(mod_iiv, events = maint_only[maint_only$arm == "No furosemide", ],
                  keep = c("arm"), returnType = "data.frame"),
  rxode2::rxSolve(mod_iiv, events = maint_only[maint_only$arm == "Furosemide", ],
                  keep = c("arm"), returnType = "data.frame")
)

pta <- sim_maint |>
  dplyr::filter(!is.na(Cc), time %in% c(4, 4.5, 5)) |>
  dplyr::mutate(total_mmol = (Cc + BASELINE) / MG_MW) |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(median_mmol = median(total_mmol),
                   pct_in_window = 100 * mean(total_mmol >= 2 & total_mmol <= 3.5),
                   .groups = "drop")

pta |>
  dplyr::rename("Arm" = arm, "Time (h)" = time,
                "Median total Mg (mmol/L)" = median_mmol,
                "% in 2.0-3.5 mmol/L" = pct_in_window) |>
  knitr::kable(digits = 2,
               caption = "Maintenance-dose-only day-2 regimen. Deng 2024 observed 9.80% target attainment at 4-5 h.")
```

The model reproduces the paper's central clinical finding qualitatively: a
maintenance dose alone leaves the typical woman well below the therapeutic
window at 4-5 h, which is the observation that motivated the paper's
dose-optimisation exercise. The attained percentage is lower than the
observed 9.80% for the same reason the Figure 4 amplitude is low, so this is
reported as a directional check and not asserted as a numeric match.

```{r target-attainment-gate}
# Directional only: the typical woman must be below the 2.0 mmol/L threshold on
# a maintenance dose alone, which is the paper's qualitative conclusion.
stopifnot(all(pta$median_mmol[pta$arm == "No furosemide"] < 2))
```

# Assumptions and deviations

## 1. The normalising constant in the printed clearance equation (major)

Deng 2024 prints its final covariate model in the Results section as

```
CL(L/h) = 2.98 x (CCR/51)^0.39 x (BMI/51)^-0.54 x (1 - 0.16 x furosemide) x exp(etaCL)
V(L)    = 25.07 x (1 - 0.25 x furosemide) x exp(etaV)
```

The constant `51` is used to normalise **both** covariates, and 51 is the
number of subjects in the study (`n = 51`, Table 1 header). It is not a
plausible reference value for either covariate: creatinine clearance has a
cohort mean of 182 mL/min and BMI a median of 29 kg/m^2. This model instead
normalises by the study medians, `(CRCL/175)^0.39` and `(BMI/29)^-0.54`.

Four independent lines of evidence support that reading.

1. **Table 2 labels 2.98 as `tvCL`**, with the footnote "tv, typical
   population value". Read literally, the printed equation puts clearance at
   the study median covariates at
   `2.98 * (175/51)^0.39 * (29/51)^-0.54 = 6.54 L/h`, so 2.98 would be a bare
   scaling constant and not the typical value the table says it is.
2. **The Abstract and Discussion both state the population value directly**:
   "The population parameter estimates were as follows: CL was 2.98 L/h" and
   "The model estimate of the population CL was 2.98 L/h".
3. **The Discussion's literature comparison only works at 2.98.** It says
   "Our study found the CL value of 2.98 CL/h, which is lower than the values
   typically described in literature", against a quoted literature range of
   1.38-5.00 L/h. A clearance of 6.54 L/h would be *above* that range, which
   would invert the sentence.
4. **The paper's own Figure 4 agrees on shape only under this reading.** The
   ratio of modelled to published drug-attributable concentration across
   t = 1-7 h is flat under the implemented reading (mean 0.659, CV 1.0%
   without furosemide; mean 0.633, CV 1.6% with), meaning the two profiles
   differ by a constant factor and therefore share the same `CL/V`. Under the
   literal reading the same ratio drifts monotonically from 0.61 to 0.40
   (CV 15% and 14%). Note that this argument establishes only that `CL/V`
   is right; it says nothing about the absolute scale, which Figure 4 gets
   wrong under either reading (see deviation 3).

The two reference values themselves come from the paper's Monte Carlo
section, which states them verbatim: "when CCR and BMI are taken at the
median (CCR:175 ml/min, BMI:29 kg/m2)". Table 1 reports the BMI median as
29.13 rather than 29; using 29.13 changes the BMI multiplier by 0.24% and is
immaterial. Table 1 gives no CCR median, only the mean 182.18 +/- 67.15;
using 182.18 in place of 175 changes the CCR multiplier by 1.7%.

**Consequence for users.** Anyone re-deriving this model from the printed
equation alone will obtain clearances 2.2-fold higher than this
implementation. The exponents, the furosemide terms, `V` and the variance
terms are unaffected.

## 2. No endogenous magnesium baseline (major)

Deng 2024 Table 2 estimates no baseline magnesium parameter, so the packaged
model predicts **administered** magnesium only and `Cc` goes to zero after
washout. The data the model was fit to are total serum magnesium: Table 1
reports an observed baseline of 0.76 mmol/L (18.5 mg/L) and Figure 3 shows
observations spanning 20-52 mg/L, none of which approach zero.

This is a genuine misspecification in the source publication rather than an
extraction choice, and the paper appears to have worked around it in
simulation without recording it as a parameter: every curve in Figure 4
starts at 0.80 mmol/L, which can only come from a baseline term added after
the fact. The paper also notes that its conditional weighted residuals
"still had an obvious trend, suggesting that there may be important factors
that are not found to affect the accuracy of the model" -- a missing baseline
is exactly the kind of factor that produces such a trend.

The model is packaged as published. This vignette adds the reported
0.76 mmol/L baseline explicitly wherever it compares against a published
total-magnesium quantity, and users comparing against measured total serum
magnesium should add roughly 18.5 mg/L to `Cc`. Both sibling models estimate
a baseline directly (`Salinger_2013_magnesiumSulfate` BL = 20.8 mg/L,
`Easterling_2018_magnesium_sulfate` BL = 22.48 mg/L), so a user moving
between the three models must handle this difference deliberately.

## 3. Figure 4 amplitude is not reproducible (erratum)

With the baseline added, the replication above tracks the *shape* of Figure 4
closely but sits below it by a near-constant factor of about 0.66 (no
furosemide) or 0.63 (furosemide) on the drug-attributable concentration.
Because that factor varies by only 1-2% across seven hourly time points in
both arms, the disagreement is an amplitude effect -- a dose-amount or volume
scaling -- and not a structural one: a wrong `CL/V` would make the ratio drift
with time, as the shape test above shows it does under the alternative
reading.

Reproducing Figure 4's amplitude with `V = 25.07 L` would require the 10 g
maintenance dose to deliver about 1.53 g of elemental magnesium, whereas
10 g of MgSO4-7H2O delivers 0.986 g and 10 g of the anhydrous salt would
deliver 2.02 g; neither matches, and no pharmaceutical hydrate of magnesium
sulfate has the implied molecular weight. The discrepancy is therefore
reported and left uncorrected. **No parameter has been tuned to close it.**
The Table 2 estimates, which are the reusable content of the paper, reproduce
their own closed-form exposure metrics exactly (gates 1 and 2 and the PKNCA
comparison above).

## 4. Interpretation of the variance terms

Table 2's inter-individual variability rows are labelled `omega^2 V` and
`omega^2 CL` and are read here as log-scale **variances** (0.023 and 0.082,
i.e. CV 15.3% and 29.2%). The header of the adjacent column reads "CV (%)"
and gives 26.89 for V and 17.38 for CL, but those cannot be the IIV CVs: the
smaller variance would then have to carry the larger CV. They are the
relative standard errors of the variance estimates, consistent with the
bootstrap column alongside them. No eta correlation is reported, so the two
random effects are coded independently.

## 5. Other assumptions

* **Salt basis.** The paper never states whether "10 g" refers to the
  heptahydrate or the anhydrous salt. The heptahydrate is assumed, matching
  obstetric practice and both sibling models.
* **Creatinine clearance assay.** The paper reports "creatinine clearance
  (CCR)" without naming an estimating equation, and does not state whether
  the values are body-surface-area normalised. They are treated as raw
  mL/min. The `CRCL` canonical entry has ample precedent for raw
  un-normalised values.
* **Covariate distributions.** Table 1 reports CRCL as a mean with SD and BMI
  as a median with IQR; the virtual cohort draws both log-normally to match
  those summaries. Body weight is never reported, so no allometric term is
  possible and none is in the model.
* **Furosemide is confounded with disease severity.** The paper attributes the
  furosemide effect to the volume overload and renal impairment that prompted
  the prescription rather than to a drug interaction. The indicator should not
  be transported to a population in which furosemide is given for another
  reason.
* **Infusion durations.** Study infusions ran 30-120 min (loading) and 6-8 h
  (maintenance) at clinician discretion. This vignette uses the 20 min and
  6.7 h durations of the paper's own Monte Carlo regimens so the replication
  is comparable to Figure 4.
* **Screened but not retained.** Age, serum albumin, calcium, labetalol,
  nifedipine, adverse-reaction occurrence, gestational age, diagnosis and
  mode of delivery were all screened by the stepwise covariate model and
  rejected; no point estimates are reported for any of them. Age and albumin
  are recorded in the model file's `covariatesDataExcluded` for provenance.
* **Calcium units.** Table 1 tabulates calcium as "2.17 +/- 0.15 g/L", which
  is almost certainly mmol/L. Calcium is not in the model, so nothing depends
  on it.
* **No external validation.** The paper states that the model was only
  internally evaluated.
