---
title: "Imipenem in critically ill adults, parametric vs nonparametric (de Velde 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Imipenem in critically ill adults, parametric vs nonparametric (de Velde 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

de Velde 2020 fitted **two** population PK models to the *same* imipenem
dataset from critically ill adults: a parametric model in NONMEM 7.2 (FOCE-I)
and a nonparametric model in Pmetrics 1.5.2 (NPAG). The two were built
independently to a pre-specified workflow and converged on the same structure:
two compartments, parameterised in micro-constants, with absolute
(BSA-unadjusted) CKD-EPI eGFR as a power covariate on the elimination rate
constant. Both are packaged here.

```{r models}
mod_par <- readModelDb("deVelde_2020_imipenem")
mod_np  <- readModelDb("deVelde_2020_imipenem_nonparametric")
```

* Parametric (NONMEM): `r rxode2::rxode(mod_par)$description`
* Nonparametric (Pmetrics NPAG): `r rxode2::rxode(mod_np)$description`
* Citation: `r rxode2::rxode(mod_par)$reference`
* Article (open access): <https://doi.org/10.1007/s40262-020-00859-1>

`deVelde_2020_imipenem` was first added to this package from the Zhang 2025
imipenem systematic review (see the
[Zhang 2025 review article](Zhang_2025_imipenem_model_review.html)). It has
since been re-verified against this primary publication, and the
nonparametric arm was added at the same time. The differences the primary
exposed are listed under "Assumptions and deviations" below.

# Population

Data came from a prospective cohort study (2010-2013) in the intensive care
unit of the Geneva University Hospitals, Switzerland. Of 54 imipenem-treated
patients, the last 27 had exact dosing and sampling times recorded; one was
excluded for missing height, leaving 26. Inclusion required suspected or
documented severe bacterial infection and age 18-60 years; exclusion criteria
were Cockcroft-Gault eGFR below 60 mL/min, BMI below 18 or above 30 kg/m^2, and
pregnancy. The cohort was 69% male, median age 51 years (IQR 39-54), median
weight 75 kg (IQR 66-85), median BSA 1.89 m^2, and median APACHE II score 22.
Renal function was high: median CKD-EPI 116 mL/min/1.73 m^2 and CKD-EPI-abs
119 mL/min at inclusion. Most infections were lower respiratory tract (62%). No
patient received continuous renal replacement therapy (Table 1, Sect. 2.1,
Sect. 3.1).

Patients received imipenem/cilastatin 500 mg/500 mg four times daily as a
30-min intravenous infusion. 138 plasma samples (peak, intermediate and trough
on days 1, 2, 3, 4 and 6) were assayed by HPLC-UV; the 13 below the 0.5 mg/L
limit of quantification were excluded, leaving 125 concentrations after 84
doses (Sect. 3.2).

```{r population-metadata}
str(mod_par()$population[c("n_subjects", "n_concentrations", "regions", "renal_function")])
```

# Source trace

The per-parameter origin is also recorded as an in-file comment next to each
`ini()` entry in both model files.

| Equation / parameter | NONMEM model | Pmetrics model | Source location |
|---|---|---|---|
| Two-compartment structure, micro-constants | yes | yes | Sect. 3.3 / 3.4 (1-cmt dOFV +22.99 / d-2LL +26.5; 3-cmt no gain) |
| `lkel` (Ke at CRCL = 119) | log(0.637) | log(0.681) | Table 2 |
| `lk12` (Kcp) | log(0.166) | log(0.374) | Table 2 |
| `lk21` (Kpc) | log(0.195) | log(0.495) | Table 2 |
| `lvc` (Vc) | log(29.6) | log(31.1) | Table 2 |
| `e_crcl_kel` (Ke(cov)) | 0.655 | 0.658 | Table 2; Eq. 9 / Eq. 10 |
| `kel = Ke x (CRCL/119)^Ke(cov) x e^eta` | yes | yes (exponent also random) | Eq. 9 / Eq. 10; 119 = median CKD-EPI-abs (Table 1) |
| `etalkel` | 0.0354 | log(0.340^2 + 1) | Sect. 3.3 (omega^2 = 0.0354); Table 2 CV 34.0% |
| `etalk12`, `etalk21`, `etalvc` | none | log(CV^2 + 1) from 81.2%, 72.0%, 42.6% | Sect. 3.3 (no BSV on Vc, Kcp, Kpc); Table 2 |
| `etae_crcl_kel` | none | log(0.552^2 + 1) | Table 2 (Ke(cov) CV 55.2%) |
| Residual error | `lnorm(expSd)`, `expSd` = 0.348 | `add(0.17) + prop(0.17) + combined1()` | Eq. 1 and Table 2; Eqs. 3 and 5, gamma 3.40 (Table 2), C0 = C1 = 0.05 (Sect. 3.4) |
| `d/dt(central)`, `d/dt(peripheral1)` | yes | yes | standard two-compartment micro-constant ODEs |

The Pmetrics residual error follows from Eqs. 3 and 5, `error = gamma x (C0 +
C1 x OBS)`, so the SD is `3.40 x 0.05 + 3.40 x 0.05 x C = 0.17 + 0.17 x C` mg/L.

# Closed-form checks against the published parameters

Neither model is fitted here, so the first validation is that the packaged
files reproduce the paper's own derived quantities exactly.

```{r closed-form}
omega_cv <- function(mod, eta) {
  om <- rxode2::rxode(mod)$omega
  sqrt(exp(om[eta, eta]) - 1)
}
typical <- function(mod) {
  th <- rxode2::rxode(mod)$theta
  kel <- exp(th[["lkel"]]); k12 <- exp(th[["lk12"]]); k21 <- exp(th[["lk21"]])
  vc  <- exp(th[["lvc"]])
  a <- kel + k12 + k21
  beta <- (a - sqrt(a^2 - 4 * k21 * kel)) / 2
  alpha <- k21 * kel / beta
  c(kel = kel, vc = vc, cl = kel * vc, vss = vc * (1 + k12 / k21),
    thalf_alpha = log(2) / alpha, thalf_beta = log(2) / beta)
}
tv <- rbind(NONMEM = typical(mod_par), Pmetrics = typical(mod_np))

checks <- tibble::tribble(
  ~Check, ~Published, ~Model,
  "NONMEM Ke CV (Table 2, Eq. 2)", 0.190, omega_cv(mod_par, "etalkel"),
  "NONMEM Ke SD, 1/h (Fig. 1)", 0.121, 0.637 * omega_cv(mod_par, "etalkel"),
  "Pmetrics Ke CV (Table 2)", 0.340, omega_cv(mod_np, "etalkel"),
  "Pmetrics Kcp CV (Table 2)", 0.812, omega_cv(mod_np, "etalk12"),
  "Pmetrics Kpc CV (Table 2)", 0.720, omega_cv(mod_np, "etalk21"),
  "Pmetrics Vc CV (Table 2)", 0.426, omega_cv(mod_np, "etalvc"),
  "Pmetrics Ke(cov) CV (Table 2)", 0.552, omega_cv(mod_np, "etae_crcl_kel")
) |>
  mutate(`Rel. diff` = signif(Model / Published - 1, 2), Model = signif(Model, 4))
knitr::kable(checks, caption = "Between-subject variability reproduced from the packaged omegas.")
# Identities of the encoded omegas: the only slack is the paper's rounding.
stopifnot(all(abs(checks$`Rel. diff`) < 0.006))

knitr::kable(signif(tv, 4), caption = paste(
  "Typical-subject (CRCL = 119 mL/min) derived quantities. CL = Vc x Ke is",
  "the clearance quoted in the Discussion."
))
stopifnot(abs(tv["NONMEM", "cl"] - 0.637 * 29.6) < 1e-9)
```

The typical NONMEM clearance, `Vc x Ke` = `r round(tv["NONMEM", "cl"], 1)` L/h,
is the value the Discussion calls higher than earlier imipenem models in
critically ill patients and attributes to augmented renal clearance. Imipenem's
disposition is dominated by the alpha phase (half-life
`r round(tv["NONMEM", "thalf_alpha"], 2)` h in the NONMEM model), consistent with
the roughly 1 h half-life the Introduction quotes for normal renal function.

# Virtual cohort and simulation

The paper's Electronic Supplementary Fig. 4 simulates each model at
CKD-EPI-abs 150, 120 and 90 mL/min. The same three renal-function levels are
used here, with 200 virtual patients per model and level, on the study regimen
of 500 mg every 6 h as a 30-min infusion for 4 days. CKD-EPI-abs is held
constant per patient (the source analysis treated it as time-varying; see
"Assumptions and deviations").

```{r cohort}
set.seed(2020)
rxode2::rxSetSeed(2020)

n_per_arm <- 200L
crcl_levels <- c(150, 120, 90)
dose_times <- seq(0, 90, by = 6)
obs_times <- sort(unique(c(seq(90, 96, by = 0.25), 90.5)))

make_cohort <- function(crcl, id_offset) {
  ids <- id_offset + seq_len(n_per_arm)
  dose <- expand.grid(id = ids, time = dose_times) |>
    mutate(evid = 1L, amt = 500, rate = 1000, cmt = "central")
  obs <- expand.grid(id = ids, time = obs_times) |>
    mutate(evid = 0L, amt = 0, rate = 0, cmt = "central")
  bind_rows(dose, obs) |>
    mutate(CRCL = crcl, egfr = paste0("CKD-EPI-abs ", crcl, " mL/min")) |>
    arrange(id, time, desc(evid))
}
events <- bind_rows(lapply(seq_along(crcl_levels), function(i) {
  make_cohort(crcl_levels[i], id_offset = (i - 1L) * n_per_arm)
}))
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

```{r simulate}
sim_one <- function(mod, label) {
  rxode2::rxSolve(mod, events = events, keep = c("egfr", "CRCL"),
                  returnType = "data.frame") |>
    mutate(model = label)
}
sim <- bind_rows(
  sim_one(mod_par, "NONMEM (parametric)"),
  sim_one(mod_np, "Pmetrics (nonparametric)")
) |>
  mutate(tad = time - 90,
         egfr = factor(egfr, levels = paste0("CKD-EPI-abs ", crcl_levels, " mL/min")))
```

# Replicate published figures

```{r figure-esm4, fig.height = 6}
# Replicates Electronic Supplementary Fig. 4 of de Velde 2020: simulated
# concentration-time profiles by CKD-EPI-abs for each model. `Cc` is the
# individual prediction (no residual error).
band <- sim |>
  group_by(model, egfr, tad) |>
  summarise(
    P025 = quantile(Cc, 0.025), P50 = median(Cc), P975 = quantile(Cc, 0.975),
    .groups = "drop"
  )
ggplot(band, aes(tad, P50, colour = model, fill = model)) +
  geom_ribbon(aes(ymin = P025, ymax = P975), alpha = 0.2, colour = NA) +
  geom_line() +
  facet_wrap(~egfr, ncol = 1) +
  scale_y_log10() +
  labs(x = "Time after dose (h), steady state", y = "Imipenem (mg/L)",
       colour = NULL, fill = NULL,
       title = "Median and 95% prediction interval, 500 mg q6h (30-min infusion)",
       caption = "Replicates Electronic Supplementary Fig. 4 of de Velde 2020.") +
  theme(legend.position = "bottom")
```

The Discussion states that "the concentrations of the 2.5th percentile are
approximately twofold lower for the nonparametric model". The table below
reads the 2.5th percentile off both models at mid-interval and at trough.

```{r p025-claim}
p025 <- band |>
  filter(tad %in% c(3, 6)) |>
  select(model, egfr, tad, P025) |>
  pivot_wider(names_from = model, values_from = P025) |>
  mutate(Ratio = `NONMEM (parametric)` / `Pmetrics (nonparametric)`)
p025 |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  rename("Time after dose (h)" = tad, "Renal function" = egfr,
         "Ratio NONMEM / Pmetrics" = Ratio) |>
  knitr::kable(caption = "2.5th percentile of simulated concentration (mg/L).")
# Gate the DIRECTION only, on the median ratio across the six cells rather
# than any one cell's tail quantile (a 2.5th percentile of 200 subjects is
# the 5th-lowest value and moves between runs). The magnitude is a known
# deviation -- see the text below -- and is deliberately not gated.
stopifnot(median(p025$Ratio) > 1.5)
```

The direction reproduces: the nonparametric model's lower tail sits well below
the parametric one at every renal-function level. The magnitude does not. The
median ratio here is `r signif(median(p025$Ratio), 2)`, against the paper's
"approximately twofold". The likely cause is the parametric approximation of
the nonparametric arm. NPAG searched strict parameter boundaries (Ke 0-1.5
1/h, V 1-70 L, Kcp and Kpc 0-1 1/h; Sect. 3.4) and placed only 16 support
points. The unbounded log-normal marginals used here, with CVs of 72-81% on
the distribution rate constants, reach further into the tails than that
discrete distribution can. The medians of the two models agree closely (see
the NCA table below); it is the tails that differ. Use the nonparametric model
for central tendency and treat its extreme percentiles with caution.

# PKNCA validation

## Typical subject: exact exposure identities

For a typical subject (random effects removed), the steady-state AUC over a
dosing interval must equal `dose / CL` and the terminal half-life must equal
the closed-form beta-phase half-life. The reference values below are computed
from the published parameters, not from the paper's text; the paper reports
no NCA.

```{r pknca-typical}
typ_events <- bind_rows(
  data.frame(id = 1L, time = dose_times, evid = 1L, amt = 500, rate = 1000,
             cmt = "central"),
  data.frame(id = 1L, time = c(seq(0, 96, by = 0.1), seq(96.5, 160, by = 0.5)),
             evid = 0L, amt = 0, rate = 0, cmt = "central")
) |>
  mutate(CRCL = 119) |>
  arrange(time, desc(evid))
# The last dose is at 90 h, so the tail from 96 h onward is a washout that
# PKNCA uses for the terminal half-life.

typ_sim <- bind_rows(
  rxode2::rxSolve(rxode2::zeroRe(mod_par), events = typ_events,
                  returnType = "data.frame", atol = 1e-10, rtol = 1e-10) |>
    mutate(treatment = "NONMEM"),
  rxode2::rxSolve(rxode2::zeroRe(mod_np), events = typ_events,
                  returnType = "data.frame", atol = 1e-10, rtol = 1e-10) |>
    mutate(treatment = "Pmetrics")
) |>
  mutate(id = 1L)

typ_conc <- typ_sim |> filter(!is.na(Cc)) |> select(id, time, Cc, treatment)
typ_dose <- bind_rows(
  typ_events |> filter(evid == 1) |> mutate(treatment = "NONMEM"),
  typ_events |> filter(evid == 1) |> mutate(treatment = "Pmetrics")
) |>
  select(id, time, amt, treatment)

typ_int <- data.frame(start = c(90, 90), end = c(96, 160),
                      auclast = c(TRUE, FALSE), half.life = c(FALSE, TRUE))
typ_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(typ_conc, Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(typ_dose, amt ~ time | treatment + id),
  intervals = typ_int
))

reference <- tibble::tibble(
  treatment = c("NONMEM", "Pmetrics"),
  auclast   = 500 / tv[, "cl"],
  half.life = tv[, "thalf_beta"]
)
cmp_typ <- nlmixr2lib::ncaComparisonTable(
  simulated = typ_nca, reference = reference, by = "treatment",
  params = c("auclast", "half.life"),
  units = c(auclast = "mg*h/L", half.life = "h"),
  tolerance_pct = 20
)
knitr::kable(cmp_typ, caption = paste(
  "Typical subject at CKD-EPI-abs 119 mL/min: PKNCA vs closed form",
  "(AUC over the steady-state interval 90-96 h = dose / CL; terminal",
  "half-life = ln 2 / beta). * marks a difference of more than 20%."
))

typ_res <- as.data.frame(typ_nca) |>
  filter(PPTESTCD %in% c("auclast", "half.life")) |>
  select(treatment, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  left_join(reference, by = "treatment", suffix = c("", "_ref"))
# Same parameters on both sides, so only numerical error separates them: the
# 90-96 h interval is 4 days into q6h dosing (>99.9% of steady state) and the
# linear-trapezoid AUC on a 0.1 h grid is within 0.5%.
stopifnot(
  all(abs(typ_res$auclast / typ_res$auclast_ref - 1) < 0.01),
  all(abs(typ_res$half.life / typ_res$half.life_ref - 1) < 0.02)
)
```

## Virtual cohort: steady-state exposure by renal function

```{r pknca-cohort}
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  mutate(treatment = paste(sub(" .*", "", model), egfr, sep = " | ")) |>
  select(id, time, Cc, treatment)

dose_df <- bind_rows(lapply(unique(sim_nca$treatment), function(tr) {
  ids <- unique(sim_nca$id[sim_nca$treatment == tr])
  events |> filter(evid == 1, id %in% ids) |> mutate(treatment = tr)
})) |>
  select(id, time, amt, treatment)

intervals <- data.frame(start = 90, end = 96, cmax = TRUE, cmin = TRUE,
                        auclast = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id),
  intervals = intervals
))

nca_summary <- as.data.frame(nca_res) |>
  group_by(treatment, PPTESTCD) |>
  summarise(
    value = sprintf("%.2f [%.2f-%.2f]", median(PPORRES),
                    quantile(PPORRES, 0.05), quantile(PPORRES, 0.95)),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = PPTESTCD, values_from = value)
nca_summary |>
  rename("Model | renal function" = treatment,
         "Cmax (mg/L)" = cmax, "Cmin (mg/L)" = cmin,
         "AUC0-6 at SS (mg*h/L)" = auclast) |>
  knitr::kable(caption = "Simulated steady-state NCA, median [5th-95th percentile], 500 mg q6h.")

# Structural check: in the NONMEM model, AUC over a steady-state interval is
# dose / CL_i with CL_i = Vc x Ke_i, so the median AUC at each renal-function
# level tracks dose / (Vc x Ke x (CRCL/119)^0.655) -- the lognormal eta has
# median 1. A mis-transcribed Ke, Vc or exponent moves this by tens of percent.
auc_par <- as.data.frame(nca_res) |>
  filter(PPTESTCD == "auclast", grepl("^NONMEM", treatment)) |>
  mutate(CRCL = as.numeric(sub(".*CKD-EPI-abs ([0-9]+).*", "\\1", treatment))) |>
  group_by(CRCL) |>
  summarise(auc_med = median(PPORRES), .groups = "drop") |>
  mutate(auc_pred = 500 / (0.637 * 29.6 * (CRCL / 119)^0.655))
knitr::kable(signif(auc_par, 4), caption = "NONMEM model: median simulated AUC0-6 vs dose / typical CL.")
stopifnot(all(abs(auc_par$auc_med / auc_par$auc_pred - 1) < 0.08))
```

# Assumptions and deviations

- **Re-verification against the primary.** `deVelde_2020_imipenem` was first
  transcribed from the Zhang 2025 systematic review. Checked against this
  primary publication, every structural value (Ke 0.637, Kcp 0.166, Kpc 0.195,
  Vc 29.6, Ke(cov) 0.655, reference 119 mL/min) and the 19% IIV on Ke agree
  exactly. Three things changed:
  - **Residual error.** The review labelled the NONMEM residual "Proportional =
    34.8%". The primary fitted log-transformed concentrations with an additive
    error on the log scale (Eq. 1; Table 2 "Exponential error"), so the model
    now uses `lnorm(expSd)` with `expSd = 0.348` instead of `prop(0.348)`. The
    two agree closely for small errors; `lnorm` is the structure the authors
    fitted.
  - **Omega.** The primary states the variance directly (`omega^2 = 0.0354`,
    Sect. 3.3), replacing the value back-calculated from the rounded 19% CV
    (0.0355).
  - **Study location.** The patients were treated in Geneva, Switzerland; the
    review listed the Netherlands, which is the authors' affiliation. The
    population metadata now also carries the 30-min infusion, the 125 analysed
    concentrations (138 drawn, 13 below LOQ) and the renal-function summary.
- **Nonparametric arm, approximated parametrically.** NPAG estimates a discrete
  joint distribution (16 support points, Fig. 1), not a parametric omega
  matrix, and the support points are not published. The Pmetrics model
  approximates that distribution with independent log-normal marginals whose
  CVs match Table 2 exactly (`omega^2 = log(CV^2 + 1)`). Correlations are not
  reported (the paper only states none exceeded 0.95), so the marginals are
  independent. The approximation is required to simulate the arm at all, and
  it is the same treatment given to the other Pmetrics models in this package
  (`Tsai_2023_ceftriaxone`, `Hughes_2024_vancomycin_nonparametric`).
- **Pmetrics typical values are means.** Table 2 reports the Pmetrics
  population by its probability-weighted mean; the model uses each mean as the
  median of its log-normal marginal, following `Tsai_2023_ceftriaxone`. The
  model's own mean therefore sits `exp(omega^2 / 2)` above the printed mean:
  6% for Ke and 29% for Kcp. The bootstrap medians in Table 2 (for example Ke
  0.586 1/h) are a different summary and were not used.
- **Pmetrics covariate exponent is random.** Eq. 10 writes the individual Ke as
  `Ke_i,med x (CKD-EPI-abs_i / 119)^Ke(cov)_i,med`: each subject has their own
  exponent. The model therefore gives `e_crcl_kel` its own log-normal eta (CV
  55.2%, Table 2), in the same way as `Downes_2023_vancomycin_full`.
- **Pmetrics residual error evaluated at the prediction.** The assay-error
  polynomial (Eq. 5) is evaluated at the observation in Pmetrics and at the
  prediction here, which is how nlmixr2 residual models work. The additive and
  proportional parts add linearly (`combined1()`), as in Eq. 3.
- **Constant renal function.** Both source models used CKD-EPI-abs as a
  time-varying covariate (a median of three creatinine samples per patient;
  NOCB in NONMEM, LOCF with interpolation in Pmetrics). The models accept a
  time-varying `CRCL` column; the simulations above hold it constant per
  patient.
- **Supply absolute eGFR.** `CRCL` here is CKD-EPI multiplied by the patient's
  BSA, in mL/min. A standard mL/min/1.73 m^2 value understates the ratio for
  any patient with a BSA above 1.73 m^2 (the cohort median was 1.89 m^2).
- **The 2.5th-percentile claim reproduces in direction, not magnitude.** The
  Discussion's "approximately twofold lower" for the nonparametric model comes
  out at roughly three- to fivefold here, because the log-normal approximation
  ignores the strict NPAG parameter boundaries. This is a known deviation of
  the approximation, not a transcription error: the medians agree, and the
  CVs match Table 2 exactly. Only the direction is gated.
- No erratum or correction notice for this article was found in Europe PMC
  (checked 2026-09-26).
