---
title: "Tideglusib (Di Deo 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tideglusib (Di Deo 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

This paper contributes **two** models, extracted as two files because the
authors fitted two separate final models on two different cohorts (the
Phase I fit was used as a Bayesian prior for the Phase II fit, not merged
with it):

* `DiDeo_2025_tideglusib_healthy` -- Phase I study NP031112-07A03, elderly
  healthy subjects (n = 54, 1832 samples, rich sampling).
* `DiDeo_2025_tideglusib_dm1` -- Phase II study AMO-02-MD-2-001, adolescent
  and adult DM-1 patients (n = 16, 51 samples, sparse sampling), estimated
  with NONMEM `$PRIOR NWPRI` using the healthy-subject model as the prior.

```{r citation, echo = FALSE, results = "asis"}
for (nm in c("DiDeo_2025_tideglusib_healthy", "DiDeo_2025_tideglusib_dm1")) {
  ui <- rxode2::rxode(readModelDb(nm))
  cat("* `", nm, "` -- ", ui$description, "\n", sep = "")
}
```

* Citation: `r rxode2::rxode(readModelDb("DiDeo_2025_tideglusib_dm1"))$reference`
* Article: <https://doi.org/10.3390/pharmaceutics17081065>
* Supplement: <https://www.mdpi.com/article/10.3390/pharmaceutics17081065/s1>

Both models share one structure: two-compartment disposition with first-order
oral absorption, allometric body-weight scaling on all four disposition
parameters, and a **dose-dependent** absorption rate constant and relative
bioavailability that switch at a 400 mg per-administration threshold.

# Population

**Phase I (NP031112-07A03).** 72 elderly subjects were randomised across six
dose groups (12 per group, 9 active : 3 placebo); the 54 who received active
tideglusib contributed 1832 plasma samples. Age 60-74 years (mean 64.3,
s.d. 3.5), body weight 50.7-98.1 kg (mean 74.5, s.d. 9.9), 24 of 54 female.
Dose groups were 300 mg b.i.d., 400 mg b.i.d., and 600, 800, 1000 and 1200 mg
q.d. as an oral suspension of formulation F06-037F, given as a single dose,
a 48 h washout, then 14 days of repeat dosing (Di Deo 2025 Section 2.1,
Section 2.2 and Table 1, first six rows).

**Phase II (AMO-02-MD-2-001).** 16 adolescent and adult patients with
congenital or juvenile-onset DM-1, aged 13.8-34.9 years (mean 20.9, s.d. 5.9),
body weight 36.8-122.6 kg (mean 63.6, s.d. 19.6), 6 of 16 female, randomised
1:1 to 400 mg or 1000 mg q.d. for 12 weeks after a 2-week single-blind placebo
run-in. Sampling was sparse -- pre-dose and 2-4 h post-dose at weeks 2 and 12 --
giving 51 evaluable samples, 9 of them below the 1 ng/mL LLOQ and all 9
pre-dose troughs (Section 2.2 and Table 1, last two rows).

The same information is available programmatically via each model's
`population` metadata (`readModelDb("<model>")()$population`).

# Source trace

Every `ini()` entry carries an in-file comment naming its source location; the
table below collects them. All parameter values come from **Di Deo 2025
Table 2**, which reports both fits side by side.

> **Table 2 column order.** The table's title reads "in healthy elderly
> subjects and DM-1 patients", but its **columns** are ordered
> AMO-02-MD-2-001 (DM-1 patients) **first**, then NP031112-07A03 (healthy
> elderly). The column headers are correct as printed; the title order is not
> the column order. This was confirmed against the paper's own precision
> claims: Section 3.1 reports the healthy fit as "%RSE < 27.4%" (fixed
> effects) and "< 41.1%" (IIV), exactly the maxima of the NP031112-07A03 RSE
> column, while the DM-1 claims ("< 23%", "< 32%") match the AMO-02-MD-2-001
> maxima (22.1 and 31.6). Independently, the Discussion's headline figures --
> "apparent systemic clearance of 341 L/h" and an apparent steady-state volume
> of "1140 L" (= V2 + V3 = 154 + 986) -- are the AMO-02-MD-2-001 column.

| Parameter | Healthy (NP031112-07A03) | DM-1 (AMO-02-MD-2-001) | Source location |
|---|---|---|---|
| `lka` | log(0.78) | log(0.767) | Table 2, "Absorption rate constant, kA (1/h)" |
| `lka_highdose` | log(0.61) | log(0.609) | Table 2, "kA (dose > 400 mg) (1/h)" |
| `lcl` | log(327) | log(341) | Table 2, "Clearance, CL (L/h)" |
| `lvc` | log(152) | log(154) | Table 2, "Volume of distribution central compartment, V2 (L)" |
| `lq` | log(66.1) | log(66.1) | Table 2, "Intercompartmental clearance, Q (L/h)" |
| `lvp` | log(1010) | log(986) | Table 2, "Volume of distribution peripheral compartment, V3 (L)" |
| `lfdepot` | fixed(log(1)) | fixed(log(1)) | Table 2, "Bioavailability (doses <= 400 mg)" = 1, FIXED |
| `lfdepot_highdose` | log(0.85) | log(0.88) | Table 2, "Bioavailability (doses > 400 mg)" |
| `e_wt_cl_q` | fixed(0.75) | fixed(0.75) | Section 2.4.1, "fixed allometric exponents on clearance (0.75)" |
| `e_wt_vc_vp` | fixed(1) | fixed(1) | Section 2.4.1, "...and volumes (1)" |
| `etalcl` | 0.18 | 0.186 | Table 2, "eta CL variance" |
| `etalvc` | 0.33 | 0.321 | Table 2, "eta V2 variance" |
| `etalq` | 0.96 | 1.15 | Table 2, "eta Q variance" |
| `etalvp` | 0.90 | 1.01 | Table 2, "eta V3 variance" |
| `etalka` | 0.06 | 0.069 | Table 2, "eta KA variance" |
| `propSd` | sqrt(0.56) | sqrt(0.54) | Table 2, "Proportional error (doses <= 400 mg)" |
| `propSd_highdose` | sqrt(0.45) | sqrt(0.46) | Table 2, "Proportional error (doses > 400 mg)" |
| Allometry `(WT/70)^exp` | n/a | n/a | Section 2.4.1, reference weight 70 kg |
| 2-compartment ODEs, first-order absorption | n/a | n/a | Figure 1 schematic; Section 3.1 |

## Two readings that had to be settled from the source

**The IIV column is a VARIANCE, not a standard deviation.** Table 2 prints
"Population Estimate (CV%)". Taking the DM-1 column, whose values are given to
three significant figures, `CV% = sqrt(variance) * 100` reproduces every
printed CV to the digit, whereas the log-normal form
`sqrt(exp(variance) - 1) * 100` does not:

```{r omega-scale}
v          <- c(etalcl = 0.186, etalq = 1.15, etalvp = 1.01,
                etalvc = 0.321, etalka = 0.069)
printed_cv <- c(43.1, 107.2, 100.5, 56.7, 26.4)      # Table 2, DM-1 column
data.frame(
  variance          = v,
  `sqrt(var)*100`   = round(sqrt(v) * 100, 1),
  `sqrt(exp(v)-1)`  = round(sqrt(exp(v) - 1) * 100, 1),
  printed_cv        = printed_cv,
  check.names       = FALSE
)
# The sqrt(variance) column matches the printed CV%; the log-normal one does not.
stopifnot(max(abs(sqrt(v) * 100 - printed_cv)) < 0.2)
```

Table 2's footnote b prints the formula as `CV = sqrt(exp(Omega^2)) x 100`,
which does **not** reproduce its own column; the arithmetic that does is
`sqrt(variance)`. nlmixr2 takes variances on the `ini()` diagonal, so the
printed values are used directly.

**The residual-error rows are variances too.** Table 2's abbreviation footnote
defines `sigma = residual variance`, and the IIV rows of the same table are
demonstrably variances (above), so the residual values are converted to the
standard deviations nlmixr2 expects: `propSd <- sqrt(0.54)`. The `sqrt()` is
kept inline in `ini()` so the printed value stays visible in the source trace.

# Virtual cohort

Original observed data are not publicly available. The cohorts below are
virtual populations whose body-weight distributions approximate the published
per-arm medians and ranges (Table 1). Each cohort is 100-200 subjects, within
the 200-per-arm cap.

```{r cohort}
# set.seed() seeds R's RNG. It does NOT seed rxode2's simulation RNG, and
# rxode2's streams are partitioned PER SOLVER THREAD -- so this cohort is
# reproducible on this machine and different on a machine with a different
# thread count. Every assertion below is written to hold for ANY cohort the
# model can produce (see known-vignette-failure-patterns.md pattern 12).
set.seed(20250216)
rxode2::rxSetSeed(20250216)

# Observation grid: dense through the absorption/peak phase (Tmax ~ 0.7 h) so
# the trapezoidal AUC and Cmax are resolved, coarser thereafter.
grid_24 <- sort(unique(c(seq(0, 4, by = 0.05), seq(4, 24, by = 0.25))))
grid_12 <- grid_24[grid_24 <= 12]

# Build one steady-state cohort. Doses run to steady state (>= 240 h of
# dosing) and observations cover only the final dosing interval, with time
# re-based so that t = 0 is the last dose.
make_cohort <- function(n, dose, ii, wt, label, model, id_offset = 0L) {
  # 720 h of dosing, not 240: IIV on Q and V3 is ~100% CV, so subjects in the
  # tail of the peripheral-distribution etas have terminal half-lives far
  # longer than the typical ~12 h and are not at steady state after 240 h.
  n_dose <- ceiling(720 / ii)
  start  <- n_dose * ii
  grid   <- if (ii == 12) grid_12 else grid_24
  dosing <- tibble(
    id = id_offset + seq_len(n), time = 0, amt = dose, evid = 1L,
    cmt = "depot", ii = ii, addl = n_dose
  )
  obs <- tidyr::crossing(
    id = id_offset + seq_len(n), time = start + grid
  ) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central", ii = 0, addl = 0L)
  bind_rows(dosing, obs) |>
    left_join(
      tibble(id = id_offset + seq_len(n), WT = wt), by = "id"
    ) |>
    mutate(
      DOSE_HIGH = as.numeric(dose > 400),
      treatment = label, model = model, dose_mg = dose, tau = ii,
      t_start = start
    ) |>
    arrange(id, time, desc(evid))
}

# Log-normal weights matched to each arm's reported median (Table 1).
lnwt <- function(n, med, sdlog) exp(rnorm(n, log(med), sdlog))

n_arm <- 100L
events <- bind_rows(
  make_cohort(n_arm,  400, 12, lnwt(n_arm, 77.1, 0.15), "Healthy 400 mg b.i.d.",
              "DiDeo_2025_tideglusib_healthy", id_offset =   0L),
  make_cohort(n_arm, 1000, 24, lnwt(n_arm, 72.5, 0.15), "Healthy 1000 mg q.d.",
              "DiDeo_2025_tideglusib_healthy", id_offset = 100L),
  make_cohort(n_arm,  400, 24, lnwt(n_arm, 60.3, 0.24), "DM-1 400 mg q.d.",
              "DiDeo_2025_tideglusib_dm1",     id_offset = 200L),
  make_cohort(n_arm, 1000, 24, lnwt(n_arm, 58.7, 0.30), "DM-1 1000 mg q.d.",
              "DiDeo_2025_tideglusib_dm1",     id_offset = 300L)
)

# Disjoint-ID guard: duplicate ids across cohorts silently merge into a single
# subject receiving the summed dose.
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
stopifnot(n_distinct(events$id) == 4L * n_arm)
```

# Simulation

Each model is solved over its own arms, then the two are combined.

```{r simulate}
solve_arms <- function(model_name) {
  ev <- events |> filter(model == model_name)
  if (nrow(ev) == 0L) return(NULL)
  rxode2::rxSolve(
    readModelDb(model_name), events = ev,
    keep = c("treatment", "model", "dose_mg", "tau", "t_start",
             "WT", "DOSE_HIGH")
  ) |>
    as.data.frame()
}

sim <- bind_rows(lapply(
  c("DiDeo_2025_tideglusib_healthy", "DiDeo_2025_tideglusib_dm1"),
  solve_arms
)) |>
  mutate(
    id   = as.integer(as.character(id)),
    t    = time - t_start,      # time since the last (steady-state) dose
    Cng  = Cc * 1000            # model is mg/L == ug/mL; paper reports ng/mL
  ) |>
  filter(t >= 0)

# The solve must produce real concentrations for every arm -- guard against a
# silently empty or all-NA result (a gate that cannot go red is worse than none).
stopifnot(nrow(sim) > 0, !all(is.na(sim$Cc)))
stopifnot(n_distinct(sim$treatment) == 4L)
stopifnot(all(tapply(sim$Cng, sim$treatment, function(x) any(x > 0))))
stopifnot(all(sim$Cng >= 0))
```

## Structural check: the dose-dependent switches fire

`DOSE_HIGH` gates three quantities at once. `ka` is returned per subject by
the solver, so the switch can be read directly off the output.

```{r switch-check}
sim |>
  group_by(treatment, DOSE_HIGH) |>
  summarise(ka_min = min(ka), ka_max = max(ka), .groups = "drop") |>
  knitr::kable(digits = 3, caption = "Absorption rate constant by dose group.")

# Typical-value ka must equal the two printed Table 2 values exactly.
ka_tv <- function(model, dose) {
  m0 <- rxode2::zeroRe(readModelDb(model))
  ev <- rxode2::et(amt = dose, cmt = "depot") |> rxode2::et(c(0, 1))
  s  <- rxode2::rxSolve(m0, ev, params = data.frame(
    WT = 70, DOSE_HIGH = as.numeric(dose > 400)), addDosing = FALSE)
  unique(as.data.frame(s)$ka)
}
stopifnot(
  abs(ka_tv("DiDeo_2025_tideglusib_dm1",      400) - 0.767) < 1e-6,
  abs(ka_tv("DiDeo_2025_tideglusib_dm1",     1000) - 0.609) < 1e-6,
  abs(ka_tv("DiDeo_2025_tideglusib_healthy",  400) - 0.78 ) < 1e-6,
  abs(ka_tv("DiDeo_2025_tideglusib_healthy", 1000) - 0.61 ) < 1e-6
)
```

## Structural check: steady-state mass balance pins the fixed bioavailability

At steady state the amount cleared over one dosing interval must equal the
absorbed dose: `CL * AUCtau = F * dose`. It is the cheapest gate that **pins
the fixed `F`**: `F` is 1 by fiat at doses <= 400 mg and 0.88 (DM-1) /
0.85 (healthy) above, and a mis-encoded switch, allometric exponent or unit
would show up here immediately.

The gate is applied to **typical-value** profiles. There both sides of the
identity use the same parameters and the only discrepancy is integration
error, so a tight bound is the correct assertion.

```{r mass-balance}
trap <- function(t, c) sum(diff(t) * (head(c, -1) + tail(c, -1)) / 2)

grid_fine <- sort(unique(c(seq(0, 4, by = 0.01), seq(4, 24, by = 0.05))))

mb_typical <- function(model, dose, ii, wt) {
  m0     <- rxode2::zeroRe(readModelDb(model))
  n_dose <- ceiling(1440 / ii)          # long run: guarantee steady state
  start  <- n_dose * ii
  g      <- grid_fine[grid_fine <= ii]
  ev     <- rxode2::et(amt = dose, cmt = "depot", ii = ii, addl = n_dose) |>
    rxode2::et(start + g)
  s <- rxode2::rxSolve(m0, ev, params = data.frame(
    WT = wt, DOSE_HIGH = as.numeric(dose > 400)), addDosing = FALSE) |>
    as.data.frame()
  s$t <- s$time - start
  s   <- s[order(s$t), ]
  fF  <- if (dose > 400) {
    if (model == "DiDeo_2025_tideglusib_dm1") 0.88 else 0.85
  } else 1
  tibble::tibble(
    model = model, dose_mg = dose, tau = ii, WT = wt,
    ratio = unique(s$cl) * trap(s$t, s$Cc) / (fF * dose)
  )
}

mb_grid <- expand.grid(
  model    = c("DiDeo_2025_tideglusib_healthy", "DiDeo_2025_tideglusib_dm1"),
  regimen  = c("400 q12h", "400 q24h", "1000 q24h", "1200 q24h"),
  WT       = c(40, 70, 100),
  stringsAsFactors = FALSE
)
mb_grid$dose <- as.numeric(sub(" .*", "", mb_grid$regimen))
mb_grid$ii   <- as.numeric(sub(".*q([0-9]+)h", "\\1", mb_grid$regimen))

mb_tv <- bind_rows(lapply(seq_len(nrow(mb_grid)), function(i) {
  mb_typical(mb_grid$model[i], mb_grid$dose[i], mb_grid$ii[i], mb_grid$WT[i])
}))

mb_tv |>
  mutate(F_used = ifelse(dose_mg > 400,
           ifelse(model == "DiDeo_2025_tideglusib_dm1", 0.88, 0.85), 1)) |>
  group_by(model, dose_mg, tau, F_used) |>
  summarise(min_ratio = min(ratio), max_ratio = max(ratio), .groups = "drop") |>
  knitr::kable(digits = 4,
    caption = "Typical-value steady-state mass balance across body weights 40/70/100 kg: CL * AUCtau / (F * dose). Exactly 1 if the ODEs, allometry, bioavailability switch and units are all consistent.")

# Deterministic identity -> tight bound is correct (numerical error only).
stopifnot(max(abs(mb_tv$ratio - 1)) < 0.005)
```

Across the simulated cohorts the same identity holds at the centre, with a
small negative tail: IIV on `Q` and `V3` is ~100% CV, so a handful of subjects
per arm have terminal half-lives long enough that even 720 h of dosing leaves
them fractionally short of steady state. That is a property of the published
variance model, not of the encoding, so the cohort check asserts on robust
quantiles rather than the extreme.

```{r mass-balance-cohort}
mass_balance <- sim |>
  group_by(id, treatment, model, dose_mg, tau) |>
  arrange(t, .by_group = TRUE) |>
  summarise(
    cleared  = first(cl) * trap(t, Cc),                    # mg cleared over tau
    absorbed = first(dose_mg) * ifelse(first(dose_mg) > 400,
                 ifelse(first(model) == "DiDeo_2025_tideglusib_dm1", 0.88, 0.85), 1),
    .groups  = "drop"
  ) |>
  mutate(ratio = cleared / absorbed)

mass_balance |>
  group_by(treatment) |>
  summarise(min_ratio = min(ratio), median_ratio = median(ratio),
            q95_abs_dev = quantile(abs(ratio - 1), 0.95),
            .groups = "drop") |>
  knitr::kable(digits = 4,
    caption = "Cohort steady-state mass balance by arm.")

# Centre: the identity must hold to a tight tolerance in the median subject.
stopifnot(max(abs(tapply(mass_balance$ratio, mass_balance$treatment,
                         median) - 1)) < 0.01)
# Envelope: robust to which subjects land in the long-half-life tail.
stopifnot(quantile(abs(mass_balance$ratio - 1), 0.95) < 0.05)
```

# Replicate published figures

## Figure 3 -- concentration-time profiles in DM-1 patients by dose group

```{r figure-3}
sim |>
  filter(model == "DiDeo_2025_tideglusib_dm1") |>
  group_by(treatment, t) |>
  summarise(
    Q05 = quantile(Cng, 0.05), Q50 = median(Cng), Q95 = quantile(Cng, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(t, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~treatment) +
  scale_y_log10() +
  labs(x = "Time after last dose (h)", y = "Tideglusib (ng/mL)",
       title = "Steady-state profiles in DM-1 patients",
       caption = "Replicates Figure 3 of Di Deo 2025 (median and 90% prediction interval).")
```

## Figure 5 -- higher exposure in lighter subjects

Section 3.4 reports that "higher exposure is expected to occur in subjects
with lower body weight". With a fixed dose and allometric clearance this is a
structural consequence, so the trend -- not a specific slope -- is the claim
to check.

```{r figure-5}
expo <- sim |>
  group_by(id, treatment, WT) |>
  arrange(t, .by_group = TRUE) |>
  summarise(auc = trap(t, Cng), .groups = "drop")

ggplot(expo, aes(WT, auc)) +
  geom_point(alpha = 0.35, size = 0.9) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 0.6) +
  facet_wrap(~treatment, scales = "free_y") +
  labs(x = "Body weight (kg)", y = "AUC over the dosing interval (ng*h/mL)",
       title = "Exposure decreases with body weight",
       caption = "Replicates Figure 5 of Di Deo 2025.")

# The claim is a NEGATIVE weight-exposure association in every arm. This is a
# structural consequence of allometric CL with a fixed dose, not a near-zero
# effect, so the sign is a safe assertion here: Spearman rho is strongly
# negative in every arm regardless of the cohort drawn.
rho <- expo |>
  group_by(treatment) |>
  summarise(rho = cor(WT, auc, method = "spearman"), .groups = "drop")
knitr::kable(rho, digits = 3, caption = "Spearman correlation of AUC with body weight.")
stopifnot(all(rho$rho < -0.3))
```

# PKNCA validation

NCA is computed over the final (steady-state) dosing interval, with time
re-based so `t = 0` is the last dose.

```{r pknca}
# Filter ONLY on !is.na(Cc): adding `t > 0` or `Cng > 0` would drop the
# time-zero row that PKNCA needs to anchor AUC.
sim_nca <- sim |>
  filter(!is.na(Cng)) |>
  select(id, time = t, Cc = Cng, treatment)

# Guarantee a time-zero record per subject (extravascular pre-dose value).
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

stopifnot(nrow(sim_nca) > 0)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- events |>
  filter(evid == 1) |>
  transmute(id, time = 0, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

# The paper reports AUC(0-12) for all four arms of Table 3, so NCA runs over
# 0-12 h. For the b.i.d. arm that is a full dosing interval; for the q.d. arms
# it is the paper's own reporting window.
intervals <- data.frame(
  start = 0, end = 12,
  cmax = TRUE, tmax = TRUE, auclast = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))

# Guard: NCA must have produced rows for all four arms (pattern 10 -- a
# lookup that matches nothing makes every downstream all() vacuously TRUE).
stopifnot(n_distinct(as.data.frame(nca_res)$treatment) == 4L)
```

## Comparison against published NCA (Table 3)

```{r nca_compare}
published <- tibble::tribble(
  ~treatment,               ~cmax,  ~tmax, ~auclast,
  "Healthy 400 mg b.i.d.",   504.7,  0.7,    949.7,
  "Healthy 1000 mg q.d.",    702.0,  0.7,   1749.8,
  "DM-1 400 mg q.d.",        513.5,  0.75,  1218.1,
  "DM-1 1000 mg q.d.",      1170.9,  0.7,   3145.7
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "treatment",
  units         = c(cmax = "ng/mL", tmax = "h", auclast = "ng*h/mL"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = "Simulated vs. published NCA (Di Deo 2025 Table 3). * differs from reference by >20%."
)
```

The **DM-1 400 mg** arm reproduces Table 3 closely. The other three arms
deviate, and the deviations are **not** in a consistent direction -- the
healthy arms run high while the DM-1 1000 mg arm runs low -- which rules out a
systematic transcription or unit error in the model. See "Assumptions and
deviations" for the diagnosis; no parameter was tuned to close these gaps.

## The reference exposure of Section 2.7 is reproduced

Section 2.7 specifies a well-defined large simulation: adults of body weight
70-75 kg receiving 1000 mg q.d. under fasting conditions, reported as
AUC(0-24) = 2560.8 (1282.5-5286.7) ng/mL*h and Cmax = 929.6 (446.4-1834.4)
ng/mL. Unlike the Table 3 arms this is a specified virtual cohort rather than
a median over 8-9 real subjects, so it is the more reliable target.

```{r ref-adult}
set.seed(4242)
n_ref  <- 200L
ref_ev <- make_cohort(n_ref, 1000, 24, runif(n_ref, 70, 75),
                      "Reference adult 70-75 kg", "DiDeo_2025_tideglusib_dm1",
                      id_offset = 900L)

ref <- rxode2::rxSolve(readModelDb("DiDeo_2025_tideglusib_dm1"),
                       events = ref_ev, keep = c("treatment", "t_start")) |>
  as.data.frame() |>
  mutate(id = as.integer(as.character(id)), t = time - t_start,
         Cng = Cc * 1000) |>
  filter(t >= 0)

ref_pk <- ref |>
  group_by(id) |>
  arrange(t, .by_group = TRUE) |>
  summarise(cmax = max(Cng), auc24 = trap(t, Cng), .groups = "drop")

ref_tab <- tibble::tibble(
  Parameter = c("AUC(0-24) (ng*h/mL)", "Cmax (ng/mL)"),
  Simulated = c(sprintf("%.1f (%.1f-%.1f)", median(ref_pk$auc24),
                        quantile(ref_pk$auc24, .05), quantile(ref_pk$auc24, .95)),
                sprintf("%.1f (%.1f-%.1f)", median(ref_pk$cmax),
                        quantile(ref_pk$cmax, .05), quantile(ref_pk$cmax, .95))),
  Published = c("2560.8 (1282.5-5286.7)", "929.6 (446.4-1834.4)"),
  `% diff (median)` = c(
    sprintf("%+.1f%%", 100 * (median(ref_pk$auc24) - 2560.8) / 2560.8),
    sprintf("%+.1f%%", 100 * (median(ref_pk$cmax) -  929.6) /  929.6))
)
knitr::kable(ref_tab, caption = "Section 2.7 reference adult exposure, median (5th-95th percentile).")

# AUC is the exposure metric the paper's dose rationale is built on, and it is
# insensitive to the unreported residual-error treatment (see deviations
# below): with and without residual error it moved only 2422 -> 2446 in
# development. Realised median % differences of -4.1 / -5.2 / -3.6 across
# thread counts; 20 leaves headroom while still breaking on a mis-transcribed
# clearance, dose or unit, which move AUC by tens of percent.
stopifnot(abs(100 * (median(ref_pk$auc24) - 2560.8) / 2560.8) < 20)
# Envelope check, robust to which subjects land in the tails.
stopifnot(median(ref_pk$auc24) > 1500, median(ref_pk$auc24) < 4000)
```

# Assumptions and deviations

- **Body-weight distributions are assumed.** The paper reports only per-arm
  median and range (Table 1), not the full distribution. Each arm's weights
  are drawn log-normally about the published median with a spread chosen to
  approximate the published range. The Section 2.7 reference cohort is the
  exception -- the paper specifies a 70-75 kg uniform range, so no assumption
  is needed there, which is why it is the primary quantitative gate.

- **Concentration units.** Both models are parameterised in the paper's own
  units (dose mg, CL L/h, V L), so `Cc` is natively mg/L (= ug/mL). The paper
  reports ng/mL, so the vignette multiplies by 1000 rather than hardcoding a
  conversion constant in `model()`.

- **Table 3 is reproduced for the DM-1 400 mg arm and deviates for the other
  three; the model was not tuned to close the gaps.** Three findings support
  leaving the model as the parameter table specifies:

  1. **Steady-state mass balance is exact** (`CL * AUCtau / (F * dose)` = 1.000
     to four decimals, above). The ODE system, allometric scaling,
     bioavailability switch and units are therefore internally consistent, and
     the encoded clearance is exactly Table 2's.
  2. **The deviations have opposite signs.** The healthy arms simulate *higher*
     AUC(0-12) than Table 3 while the DM-1 1000 mg arm simulates *lower* (see
     the comparison table above; at typical values and each arm's median weight
     the discrepancies are +19.8%, +35.0%, +1.1% and -12.3% for the four arms
     in Table 3 order). A transcription, unit or structural error would bias
     every arm the same way.
  3. **Table 3's healthy rows are internally inconsistent with Table 2.** For
     the healthy 400 mg b.i.d. arm, Table 3's own Css of 79.2 ng/mL implies
     AUC(0-24) = 1901 ng*h/mL, hence an apparent clearance of
     800 mg / 1.901 mg*h/L = 421 L/h. Table 2's healthy CL is 327 L/h, which
     is 351 L/h at that arm's median 77.1 kg. The two published tables imply
     clearances differing by ~20%. Table 2 is the parameter table and is what
     the model encodes.

  A contributing factor for all four arms is that Table 3's medians are taken
  over only 8-9 subjects carrying very large IIV (Q and V3 have 95-107% CV), so
  each published median is itself an imprecise statistic, and the median of a
  cohort is not the typical-value prediction when IIV is that large.

- **Cmax depends on an unreported simulation detail, so it is reported but not
  gated.** Section 2.5 says secondary parameters were derived by NCA from
  "simulated individual concentration vs. time profiles" without stating
  whether residual error was included. That choice moves Cmax a long way and
  brackets the published value: for the Section 2.7 reference cohort, NCA on
  the individual predictions gives a median Cmax about 19% *below* the
  published 929.6 ng/mL, while adding the model's proportional residual error
  (67% CV at this dose) gives roughly 1430 ng/mL, about 54% *above* it. AUC
  over the same two runs moved only 2422 -> 2446 ng*h/mL, i.e. barely at all,
  because a maximum is inflated by noise whereas an integral averages it out.
  AUC is therefore the robust target and carries the assertion; Cmax is shown
  in the tables without one.

- **No maturation function.** The paper extrapolates the weight-banded
  paediatric regimen of Table 5 down to 5 kg using allometry alone. The
  Discussion states this explicitly as a limitation: "as there is incomplete
  data on the primary route of metabolism in vivo, neither a maturation
  function nor an age-dependent allometric exponent were included". The models
  reproduce that choice, so predictions in young children carry the paper's own
  caveat.

- **Food and meal type are documented but not modelled.** In the Phase II study
  every dose was given after an overnight fast with food restricted for at
  least an hour afterwards, so no fed-vs-fasted contrast is identifiable.
  Sections 2.6 and 3.3 explore the timing and type of the first post-dose meal
  post hoc against derived NCA metrics in 30 records and find no effect, so no
  meal term enters `model()`. These screened-but-unretained covariates (`AGE`,
  `BMI`, `SEXF`, `FED`) are recorded in each model's `covariatesDataExcluded`
  metadata.

- **All parameter values come from the paper's Table 2**; none was derived from
  a figure, correspondence, or an upstream model. The supplement was not
  required -- it contains inclusion/exclusion criteria, bioanalytical method
  detail, and additional summary tables and figures, but no model parameters.
