---
title: "Ceftazidime and avibactam plasma + epithelial lining fluid PK (Dimelow 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ceftazidime and avibactam plasma + epithelial lining fluid PK (Dimelow 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Dimelow 2018 developed two independent population PK models -- one per drug --
each pairing a three-compartment intravenous plasma model with a *direct
response* (instantaneous-equilibrium) link to the epithelial lining fluid (ELF)
of the lung. Because the two drugs were modelled separately, they are packaged
as two model files sharing this one vignette.

```{r models}
cef <- rxode2::rxode(readModelDb("Dimelow_2018_ceftazidime"))
avi <- rxode2::rxode(readModelDb("Dimelow_2018_avibactam"))
```

* Citation: `r cef$reference`
* Article: <https://doi.org/10.1007/s40268-018-0241-0>

```{r descriptions, echo = FALSE, results = "asis"}
cat("* **Ceftazidime** (`Dimelow_2018_ceftazidime`): ", cef$description, "\n\n", sep = "")
cat("* **Avibactam** (`Dimelow_2018_avibactam`): ", avi$description, "\n", sep = "")
```

The distinguishing feature of this analysis is the *link function* between
plasma and ELF. Section 2.2.1.1 evaluated three forms for each drug:

| Link | Equation | Selected for |
|------|----------|--------------|
| Proportional | `Celf = EPR * Cp` | neither |
| Power | `Celf = EPR(1 mg/L) * Cp^POW` | avibactam (22 OFV units better than proportional, 6 better than saturable) |
| Saturable | `Celf = EMAX * Cp / (KM + Cp)` | ceftazidime (35 OFV units better than proportional, 8 better than power) |

An effect-site equilibration model was fitted for both drugs and rejected: it
gave no OFV improvement, and the estimated ELF half-lives were 13 min
(ceftazidime, Sect. 3.2) and 8 min (avibactam, Sect. 3.3). Equilibration is
therefore treated as instantaneous, and `Celf` is an **algebraic observable**
rather than an ODE compartment in both model files.

# Population

Both models were fitted to the phase I open-label study NCT01395420
(Sect. 2.1). Forty-three healthy **male** volunteers received either
ceftazidime 2000 mg + avibactam 500 mg (cohort A, n = 22) or ceftazidime
3000 mg + avibactam 1000 mg (cohort B, n = 21) as a 2 h intravenous infusion
every 8 h for 3 days -- nine doses per subject. PK data were available for 42
subjects (22 cohort A, 20 cohort B).

Each subject contributed exactly **one** bronchoalveolar-lavage ELF sample,
which is the reason the ELF residual error is not separately identifiable
(Sect. 2.2, Sect. 4).

The paper reports no age, weight, sex-balance or race distribution table and
screened no covariates, so neither model file carries a `covariateData` entry.

```{r population}
str(cef$population)
```

# Source trace

Every `ini()` entry in both model files carries an in-file comment pointing at
its source location. They are collected here for review.

| Parameter | Ceftazidime | Avibactam | Source location |
|-----------|-------------|-----------|-----------------|
| `lcl` (CL, L/h) | 6.55 (RSE 2%) | 12.5 (RSE 1%) | Table 1, `CL` |
| `lvc` (V1, L) | 10.32 (RSE 8%) | 15.10 (RSE 10%) | Table 1, `V 1` |
| `lvp` (V2, L) | 5.82 (RSE 9%) | 6.52 (RSE 15%) | Table 1, `V 2` |
| `lvp2` (V3, L) | 0.64 (RSE 8%) | 1.58 (RSE 5%) | Table 1, `V 3` |
| `lq` (Q12, L/h) | 6.87 (RSE 24%) | 5.43 (RSE 42%) | Table 1, `Q 12` |
| `lq2` (Q13, L/h) | 0.040 (RSE 33%) | 0.14 (RSE 41%) | Table 1, `Q 13` |
| `propSd` (RESMp) | 0.101 (RSE 9%) | 0.117 (RSE 7%) | Table 1, `RESM p` |
| `lemax_elf` (EMAX, mg/L) | 45.4 (RSE 12%) | n/a | Table 2, ceftazidime `E MAX` |
| `lkm_elf` (KM, mg/L) | 71.7 (RSE 22%) | n/a | Table 2, ceftazidime `K M` |
| `lrelf` (EPR at 1 mg/L) | n/a | 0.472 (RSE 11%) | Table 2, avibactam `EPR (1 mg/l)` |
| `lpow_elf` (POW) | n/a | 0.860 (RSE 5%) | Table 2, avibactam `POW` |
| `propSd_Celf` (RESM_ELF) | 0.101, fixed | 0.117, fixed | Table 2, `RESM ELF`; Sect. 2.2 states it was fixed to the plasma median |
| BSV on CL / V1 / V2 / V3 / Q12 / Q13 | CV 10 / 18 / 18 / 25 / 66 / 61% | CV 7 / 9 / 25 / 20 / 60 / 75% | Table 1, `CV %` column |
| BSV on RESMp | CV 54% | CV 45% | Table 1, `RESM p` row, `CV %` column |
| BSV on ELF link parameters | EMAX CV 24%, KM CV 97% | EPR CV 68%, POW CV 23% | Table 2, `CV %` column |
| Plasma ODE system (3-compartment, CL/V/Q) | n/a | n/a | Sect. 2.2 ("coded as a series of differential equations") |
| `Celf` saturable link | n/a | n/a | Sect. 2.2.1.1, `C ELF (t) = E MAX . C p (t) / (K M + C p (t))` |
| `Celf` power link | n/a | n/a | Sect. 2.2.1.1, `C ELF (t) = EPR(1 mg/L) . [C p (t)]^POW` |

Between-subject variability is reported in Table 1 / Table 2 as a **%CV** of a
log-normal distribution (Sect. 2.2). The model files therefore carry
`log(1 + CV^2)` as the variance, with the arithmetic left inline so the
published CV stays legible. The conversion is confirmed by the tables'
own 5th/95th percentile columns:

```{r cv-check}
# Table 1, ceftazidime V1: median 10.32 L, CV 18%, published 5th-95th 7.73-13.78
om <- sqrt(log(1 + 0.18^2))
round(10.32 * exp(qnorm(c(0.05, 0.95)) * om), 2)

# Table 2, ceftazidime KM: median 71.7 mg/L, CV 97%, published 5th-95th 18.8-273.3
om_km <- sqrt(log(1 + 0.97^2))
round(71.7 * exp(qnorm(c(0.05, 0.95)) * om_km), 1)
```

# Plasma-ELF link functions reproduce the paper's printed penetration ratios

Sects. 3.2 and 3.3 quote six derived quantities. These are **deterministic**
functions of the Table 2 estimates -- no simulation, no cohort draw -- so they
are checked to tight tolerance.

```{r link-check}
link_cef <- function(cp) 45.4 * cp / (71.7 + cp)          # Sect. 2.2.1.1 saturable
link_avi <- function(cp) 0.472 * cp^0.860                  # Sect. 2.2.1.1 power

link_claims <- tibble::tribble(
  ~Drug,         ~Claim,                                        ~Published, ~Model,
  "Ceftazidime", "ELF:plasma ratio as Cp -> 0 (EMAX/KM, %)",     63.3,  100 * 45.4 / 71.7,
  "Ceftazidime", "Celf at Cp = 70 mg/L (mg/L)",                  22.5,  link_cef(70),
  "Ceftazidime", "Penetration at Cp = 70 mg/L (%)",              32.1,  100 * link_cef(70) / 70,
  "Ceftazidime", "Cp giving Celf = 8 mg/L -> Celf there (mg/L)",  8.0,  link_cef(15.3),
  "Ceftazidime", "Penetration at Cp = 15.3 mg/L (%)",            52.0,  100 * link_cef(15.3) / 15.3,
  "Avibactam",   "Penetration at Cp = 1 mg/L (%)",               47.0,  100 * link_avi(1),
  "Avibactam",   "Celf at Cp = 2.4 mg/L (mg/L)",                  1.0,  link_avi(2.4),
  "Avibactam",   "Penetration at Cp = 2.4 mg/L (%)",             42.0,  100 * link_avi(2.4) / 2.4,
  "Avibactam",   "Celf at Cp = 12 mg/L (mg/L)",                   4.0,  link_avi(12),
  "Avibactam",   "Penetration at Cp = 12 mg/L (%)",              33.2,  100 * link_avi(12) / 12
) |>
  mutate(`Difference (%)` = 100 * (Model - Published) / Published)

link_claims |>
  mutate(across(c(Published, Model, `Difference (%)`), \(x) round(x, 2))) |>
  knitr::kable(caption = "Sects. 3.2 / 3.3 derived quantities vs. the packaged link functions.")

# These are algebra on the published Table 2 estimates, not a simulated cohort,
# so the tolerance is tight. The residual 0-1% is the paper's own rounding of
# its printed intermediates (e.g. 22.5 for an exact 22.43).
stopifnot(max(abs(link_claims$`Difference (%)`)) < 1.5)
```

The model *files* must give the same numbers as the hand-written link functions
above. Checked against a typical-value solve so the identity is exact:

```{r link-in-model}
solve_typical <- function(mod, amt) {
  ev <- rxode2::et(amt = amt, dur = 2, cmt = "central") |>
    rxode2::et(seq(0, 8, by = 0.02), cmt = "Cc")
  as.data.frame(rxode2::rxSolve(rxode2::zeroRe(mod), ev, useLinCmt = FALSE))
}
tv_cef <- solve_typical(cef, 2000)
tv_avi <- solve_typical(avi, 500)

# Celf computed inside the model must equal the published link function applied
# to the model's own Cc, at every time point.
stopifnot(
  max(abs(tv_cef$Celf - link_cef(tv_cef$Cc))) < 1e-8,
  max(abs(tv_avi$Celf - link_avi(tv_avi$Cc))) < 1e-8
)
```

# Virtual cohort

Original observed data are not publicly available. The cohort below mirrors the
NCT01395420 design (Sect. 2.1): two dose groups, a 2 h intravenous infusion
every 8 h, nine doses. 150 subjects per arm -- below the 200/arm cap, and ample
for the percentile bands the paper plots.

```{r cohort}
# rxode2's RNG streams are partitioned per solver thread, so this cohort is
# reproducible on a given machine but NOT across thread counts. Every assertion
# below is written to hold for any cohort these models can produce.
rxode2::rxSetSeed(20180727)

n_per_arm <- 150L
tau <- 8            # dosing interval (h)
n_doses <- 9        # Sect. 2.1: nine doses per subject
t_last <- tau * (n_doses - 1)   # 64 h: start of the final infusion

# Observe densely over the FIRST and the FINAL dosing interval: the first gives
# a single-dose NCA comparison, the final gives the steady-state one the paper
# simulated (Fig. 1 is plotted "versus time from start of infusion of the last
# dose").
obs_grid <- sort(unique(c(seq(0, tau, by = 0.1), seq(t_last, t_last + tau, by = 0.1))))

make_events <- function(amt, n, id_offset, cohort) {
  rxode2::et(amt = amt, dur = 2, ii = tau, addl = n_doses - 1L, cmt = "central") |>
    rxode2::et(obs_grid, cmt = "Cc") |>
    rxode2::et(id = id_offset + seq_len(n)) |>
    as.data.frame() |>
    dplyr::mutate(cohort = cohort)
}

ev_cef <- dplyr::bind_rows(
  make_events(2000, n_per_arm, 0L,           "A: ceftazidime 2000 mg"),
  make_events(3000, n_per_arm, n_per_arm,    "B: ceftazidime 3000 mg")
)
ev_avi <- dplyr::bind_rows(
  make_events(500,  n_per_arm, 0L,           "A: avibactam 500 mg"),
  make_events(1000, n_per_arm, n_per_arm,    "B: avibactam 1000 mg")
)

# Disjoint IDs across arms are mandatory -- duplicated IDs silently merge into
# one subject receiving the summed dose. Both checks are on the RAW frame: a
# `unique()` before `anyDuplicated()` would remove the very duplicates being
# looked for and give a gate that can never go red.
stopifnot(
  anyDuplicated(ev_cef[, c("id", "time", "evid")]) == 0L,
  anyDuplicated(ev_avi[, c("id", "time", "evid")]) == 0L,
  dplyr::n_distinct(ev_cef$id) == 2L * n_per_arm,
  dplyr::n_distinct(ev_avi$id) == 2L * n_per_arm
)
```

# Simulation

Two rxode2 details are load-bearing for these models.

**`useLinCmt = FALSE` is required.** The default ODE-to-`linCmt()` conversion
silently drops the second peripheral compartment of a `k12`/`k21`/`k13`/`k31`
parameterisation and mis-maps the two endpoints onto compartment slots.

**Observation rows use `cmt = "Cc"`, not `cmt = "central"`.** Both models declare
*two* endpoints (`Cc ~ prop(...)` and `Celf ~ prop(...)`), so rxode2 builds a
`dvid`-to-`cmt` map onto the endpoint slots that follow the three ODE states;
an observation row pointing at `central` is then rejected outright. Because
`Cc` and `Celf` are already declared endpoints, naming one on an observation row
appends nothing and renumbers nothing -- the usual hazard of writing an
algebraic observable into `cmt` does not apply here. The assertion below pins
that: the ODE states must still be the three named compartments in order, and
`peripheral2` must actually carry drug.

```{r simulate}
sim_cef <- rxode2::rxSolve(cef, events = ev_cef, keep = "cohort", useLinCmt = FALSE) |>
  as.data.frame()
sim_avi <- rxode2::rxSolve(avi, events = ev_avi, keep = "cohort", useLinCmt = FALSE) |>
  as.data.frame()

stopifnot(
  # No compartment renumbering: the states, in order, are exactly the three
  # declared by d/dt(), and the deep compartment is populated (it would be
  # absent or identically zero if slot 3 had been displaced).
  identical(cef$state, c("central", "peripheral1", "peripheral2")),
  identical(avi$state, c("central", "peripheral1", "peripheral2")),
  max(sim_cef$peripheral2, na.rm = TRUE) > 0,
  max(sim_avi$peripheral2, na.rm = TRUE) > 0,
  all(c("Cc", "Celf") %in% names(sim_cef)),
  dplyr::n_distinct(sim_cef$id) == 2L * n_per_arm,
  all(sim_cef$Cc >= 0), all(sim_avi$Cc >= 0)
)
```

# Replicate published figures

## Figure 2 -- simulated ELF versus plasma concentration

Figure 2 plots simulated ELF against plasma concentration for 1000 subjects
receiving ceftazidime/avibactam 2000-500 mg q8h, with the population median and
the 5th / 95th percentiles. The ceftazidime panel is visibly saturable above
about 100 mg/L; the avibactam panel departs from linearity only mildly, above
about 20 mg/L (Sect. 3.5).

```{r figure-2, fig.height = 4}
# Replicates Figure 2 of Dimelow 2018 (cohort A dose only, as in the paper).
fig2 <- dplyr::bind_rows(
  sim_cef |> filter(cohort == "A: ceftazidime 2000 mg") |> mutate(drug = "a  Ceftazidime"),
  sim_avi |> filter(cohort == "A: avibactam 500 mg")    |> mutate(drug = "b  Avibactam")
) |>
  filter(!is.na(Cc), Cc > 0) |>
  group_by(drug) |>
  mutate(cp_bin = cut(Cc, breaks = 10^seq(-2, 2.6, by = 0.1), labels = FALSE)) |>
  group_by(drug, cp_bin) |>
  summarise(
    Cp = median(Cc),
    Q05 = quantile(Celf, 0.05), Q50 = quantile(Celf, 0.50), Q95 = quantile(Celf, 0.95),
    .groups = "drop"
  )

ggplot(fig2, aes(Cp, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~drug, scales = "free") +
  scale_x_log10() + scale_y_log10() +
  labs(
    x = "Total plasma concentration (mg/L)", y = "Total ELF concentration (mg/L)",
    title = "Figure 2 -- simulated ELF vs. plasma concentration",
    caption = "Replicates Figure 2 of Dimelow 2018 (2000-500 mg q8h). Line = median, band = 5th-95th percentile."
  )
```

## Figure 3 -- simulated ELF concentration-time profiles vs. the PK/PD targets

Figure 3 overlays the simulated ELF profiles on the plasma-derived PK/PD
targets (50% *f*T > 8 mg/L for ceftazidime, 50% *f*T > 1 mg/L for avibactam)
with a vertical line at the 4 h midpoint of the 8 h dosing interval.

```{r figure-3, fig.height = 4}
# Replicates Figure 3 of Dimelow 2018: final dosing interval, cohort A dose.
targets <- c("a  Ceftazidime" = 8, "b  Avibactam" = 1)

fig3 <- dplyr::bind_rows(
  sim_cef |> filter(cohort == "A: ceftazidime 2000 mg") |> mutate(drug = "a  Ceftazidime"),
  sim_avi |> filter(cohort == "A: avibactam 500 mg")    |> mutate(drug = "b  Avibactam")
) |>
  filter(time >= t_last, !is.na(Celf)) |>
  mutate(tad = time - t_last) |>
  group_by(drug, tad) |>
  summarise(
    Q05 = quantile(Celf, 0.05), Q50 = quantile(Celf, 0.50), Q95 = quantile(Celf, 0.95),
    .groups = "drop"
  ) |>
  mutate(target = targets[drug])

ggplot(fig3, aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2) +
  geom_line(linewidth = 0.8) +
  geom_hline(aes(yintercept = target), linetype = "dashed", colour = "firebrick") +
  geom_vline(xintercept = 4, linetype = "dotted") +
  facet_wrap(~drug, scales = "free_y") +
  scale_y_log10() +
  labs(
    x = "Time from start of the final infusion (h)", y = "Total ELF concentration (mg/L)",
    title = "Figure 3 -- simulated ELF profiles vs. plasma PK/PD targets",
    caption = paste(
      "Replicates Figure 3 of Dimelow 2018 (2000-500 mg q8h). Dashed line = PK/PD target,",
      "dotted line = 4 h dosing-interval midpoint."
    )
  )
```

## Section 3.5 -- ELF concentration at the dosing-interval midpoint

Section 3.5 reports the one fully quantitative simulation result in the paper:
at 4 h post dose, median (5th-95th percentile) simulated ELF concentrations were
**14.7 (7.1-23.4) mg/L** for ceftazidime and **1.4 (0.75-2.85) mg/L** for
avibactam, for 1000 subjects on 2000-500 mg q8h.

```{r sect-3-5}
midpoint <- function(sim, arm) {
  x <- sim |> filter(cohort == arm, abs(time - (t_last + 4)) < 1e-6)
  c(median = median(x$Celf), p05 = unname(quantile(x$Celf, 0.05)),
    p95 = unname(quantile(x$Celf, 0.95)), n = nrow(x))
}
mid_cef <- midpoint(sim_cef, "A: ceftazidime 2000 mg")
mid_avi <- midpoint(sim_avi, "A: avibactam 500 mg")
stopifnot(mid_cef[["n"]] == n_per_arm, mid_avi[["n"]] == n_per_arm)  # the gate had rows to test

tibble::tibble(
  Drug = c("Ceftazidime", "Avibactam"),
  `Published median` = c(14.7, 1.4),
  `Model median` = c(mid_cef[["median"]], mid_avi[["median"]]),
  `Published 5th-95th` = c("7.1-23.4", "0.75-2.85"),
  `Model 5th-95th` = sprintf(
    "%.2f-%.2f",
    c(mid_cef[["p05"]], mid_avi[["p05"]]), c(mid_cef[["p95"]], mid_avi[["p95"]])
  )
) |>
  mutate(`Difference in median (%)` = round(100 * (`Model median` - `Published median`) / `Published median`, 1),
         `Model median` = round(`Model median`, 2)) |>
  knitr::kable(caption = "Sect. 3.5, ELF concentration 4 h after the dose at steady state.")

# The MEDIAN is the structural quantity: a mis-transcribed clearance, volume,
# dose or link parameter moves it by tens of percent. It is checked; the
# 5th-95th spread deliberately is NOT (see Assumptions -- the published
# plasma OMEGA is a full covariance matrix whose off-diagonals were not
# printed, so the packaged diagonal block over-disperses the tails).
#
# Bounds are set OUTSIDE the range seen across two independent cohort draws,
# not tightened onto one run: ceftazidime realised -0.2% and +5.9%, avibactam
# +6.6% and -3.6%. Do not tighten these back -- rxode2 draws a different
# cohort on a machine with a different solver-thread count.
stopifnot(
  abs(mid_cef[["median"]] - 14.7) / 14.7 < 0.20,
  abs(mid_avi[["median"]] -  1.4) /  1.4 < 0.25
)
```

Section 3.5 also claims that "most subjects still achieved ceftazidime and
avibactam ELF exposures exceeding their respective plasma PK/PD targets beyond
the midpoint of the dosing interval."

```{r target-attainment}
pct_over <- function(sim, arm, target) {
  x <- sim |> filter(cohort == arm, abs(time - (t_last + 4)) < 1e-6)
  stopifnot(nrow(x) == n_per_arm)
  100 * mean(x$Celf > target)
}
attain <- c(
  Ceftazidime = pct_over(sim_cef, "A: ceftazidime 2000 mg", 8),
  Avibactam   = pct_over(sim_avi, "A: avibactam 500 mg",    1)
)
round(attain, 1)

# "Most subjects" = a majority, which is the paper's own claim and therefore
# the bound. Realised across two draws: ceftazidime 80-84%, avibactam 66-71%.
# With n = 150 the binomial SE is about 4 points, so the 50% floor sits several
# SE below either arm, and the gate still goes red if a link or disposition
# parameter is wrong (halving the penetration ratio would drop ceftazidime
# below 50%).
stopifnot(all(attain > 50))
```

# PKNCA validation

NCA is computed with PKNCA over two intervals per subject: the **first** dosing
interval (0-8 h) and the **final**, steady-state interval (64-72 h). Both
outputs are run separately -- plasma `Cc` and ELF `Celf`.

```{r pknca}
tau_intervals <- data.frame(
  start   = c(0, t_last),
  end     = c(tau, t_last + tau),
  cmax    = TRUE,
  tmax    = TRUE,
  cmin    = TRUE,
  auclast = TRUE,
  cav     = TRUE
)

dose_frame <- function(ev) {
  ev |>
    filter(evid != 0) |>
    distinct(id, cohort) |>
    tidyr::crossing(time = tau * seq(0, n_doses - 1L)) |>
    mutate(amt = 1) |>          # amount is irrelevant to the parameters requested
    arrange(id, time)
}

run_nca <- function(sim, ev, col) {
  conc <- sim |>
    dplyr::filter(!is.na(.data[[col]])) |>       # ONLY !is.na -- never time > 0 / conc > 0
    dplyr::select(id, cohort, time, conc = dplyr::all_of(col))
  stopifnot(nrow(conc) > 0, any(conc$time == 0), any(conc$time == t_last))
  co <- PKNCA::PKNCAconc(conc, conc ~ time | cohort + id)
  do <- PKNCA::PKNCAdose(as.data.frame(dose_frame(ev)), amt ~ time | cohort + id)
  suppressWarnings(PKNCA::pk.nca(PKNCA::PKNCAdata(co, do, intervals = tau_intervals)))
}

nca_cef_plasma <- run_nca(sim_cef, ev_cef, "Cc")
nca_cef_elf    <- run_nca(sim_cef, ev_cef, "Celf")
nca_avi_plasma <- run_nca(sim_avi, ev_avi, "Cc")
nca_avi_elf    <- run_nca(sim_avi, ev_avi, "Celf")
```

```{r pknca-summary}
tidy_nca <- function(res, drug, matrix) {
  as.data.frame(res$result) |>
    mutate(drug = drug, matrix = matrix,
           interval = ifelse(start == 0, "First dose (0-8 h)", "Steady state (64-72 h)"))
}
nca_all <- dplyr::bind_rows(
  tidy_nca(nca_cef_plasma, "Ceftazidime", "Plasma"),
  tidy_nca(nca_cef_elf,    "Ceftazidime", "ELF"),
  tidy_nca(nca_avi_plasma, "Avibactam",   "Plasma"),
  tidy_nca(nca_avi_elf,    "Avibactam",   "ELF")
)

nca_all |>
  filter(PPTESTCD %in% c("cmax", "tmax", "cmin", "auclast", "cav")) |>
  group_by(drug, matrix, cohort, interval, PPTESTCD) |>
  summarise(value = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  mutate(PPTESTCD = nlmixr2lib::ncaParamLabel(PPTESTCD)) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = value) |>
  mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
  dplyr::rename("Drug" = drug, "Matrix" = matrix, "Arm" = cohort, "Interval" = interval) |>
  knitr::kable(caption = "Median per-subject NCA of the simulated cohort.")
```

## ELF:plasma AUC ratio vs. the published non-compartmental result

The phase I study's own non-compartmental analysis -- the comparator this paper
was written to improve on -- found the ELF AUC to be **31-35%** of the plasma
AUC for each drug (Key Points; Sect. 4 quotes about 32% for ceftazidime and
about 35% for avibactam). That ratio is the single best end-to-end check on the
packaged models: it exercises the plasma disposition and the link function
together, over a whole dosing interval.

```{r auc-ratio}
auc_ratio <- nca_all |>
  filter(PPTESTCD == "auclast", interval == "Steady state (64-72 h)") |>
  select(drug, matrix, cohort, id, PPORRES) |>
  tidyr::pivot_wider(names_from = matrix, values_from = PPORRES) |>
  mutate(ratio = 100 * ELF / Plasma)

auc_summary <- auc_ratio |>
  group_by(drug, cohort) |>
  summarise(`Median ELF:plasma AUC ratio (%)` = round(median(ratio), 1), .groups = "drop") |>
  mutate(`Published NCA ratio (%)` = "31-35")

auc_summary |>
  dplyr::rename("Drug" = drug, "Arm" = cohort) |>
  knitr::kable(caption = "Steady-state AUC0-tau ratio, ELF vs. plasma, against the phase I NCA result.")

# The published band is 31-35%; the gate allows 25-45% so that it tolerates the
# cohort draw while still going red on a mis-transcribed EMAX, KM, EPR, POW, CL
# or V1. Realised across two independent draws of the four arms: 31.3-38.2%.
# Only the MEDIAN is gated -- the per-subject spread is inflated by the dropped
# OMEGA off-diagonals, so its tails are not a reproducible quantity.
stopifnot(all(auc_summary$`Median ELF:plasma AUC ratio (%)` > 25),
          all(auc_summary$`Median ELF:plasma AUC ratio (%)` < 45))
```

Both drugs sit inside the published band, and the ratio falls from cohort A to
cohort B for each drug -- the paper's central finding that penetration is
concentration-dependent and therefore *higher* at the lower, efficacy-relevant
plasma concentrations.

## Comparison against published NCA

The paper reports no NCA table of its own. It does quote two literature plasma
`Cmax` values while interpreting the link functions (Sects. 3.2 and 3.3,
reference [26]): about 70 mg/L for ceftazidime 2000 mg and about 12 mg/L for
avibactam 500 mg. Those are compared here against the first-dose simulated
`Cmax` of the packaged models.

```{r nca_compare}
sim_long <- nca_all |>
  filter(matrix == "Plasma", interval == "First dose (0-8 h)",
         cohort %in% c("A: ceftazidime 2000 mg", "A: avibactam 500 mg")) |>
  mutate(drug = ifelse(drug == "Ceftazidime", "Ceftazidime 2000 mg", "Avibactam 500 mg")) |>
  select(drug, PPTESTCD, PPORRES)

published <- tibble::tribble(
  ~drug,                  ~cmax,
  "Ceftazidime 2000 mg",  70,
  "Avibactam 500 mg",     12
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = sim_long,
  reference     = published,
  by            = "drug",
  params        = "cmax",
  units         = c(cmax = "mg/L"),
  tolerance_pct = 20
)
cmp |>
  dplyr::rename("Drug" = drug) |>
  knitr::kable(caption = "Simulated first-dose plasma Cmax vs. the literature values quoted by Dimelow 2018. * differs by >20%.")
```

The ceftazidime row is starred and is **not** tuned away. The discrepancy is
internal to the paper: Table 1 gives V1 = 10.32 L and CL = 6.55 L/h, which for a
2000 mg 2 h infusion imply an end-of-infusion concentration near 90 mg/L, while
Sect. 3.2 quotes about 70 mg/L from an external reference to argue that KM
(71.7 mg/L) is "comparable to the typical Cmax". Both statements are in the
source; only the Table 1 parameters are the model, so the packaged model
follows Table 1. The avibactam row agrees within tolerance.

# Assumptions and deviations

- **Plasma OMEGA is carried as diagonal.** Sect. 3.1 states that between-subject
  variability "was adequately characterized using a full covariance matrix", but
  no off-diagonal element is printed anywhere in the paper, and inventing one is
  not an option. Each parameter's published %CV is therefore carried on the
  diagonal and all correlations are set to zero. The consequence is visible and
  one-directional: medians are unaffected (they are reproduced to within 6% in
  the Sect. 3.5 check above), but the simulated 5th-95th percentile band is
  wider than the published one -- 5.2-33.1 mg/L against the paper's 7.1-23.4 for
  ceftazidime. Every assertion in this vignette is therefore written on medians,
  never on the tails.
- **Between-subject variability on the residual error is encoded, and is
  unusual.** Table 1 reports `RESMp` as a population *median* (0.101 ceftazidime,
  0.117 avibactam) with its own %CV (54% and 45%), and its 5th/95th percentile
  columns confirm a log-normal spread. The models therefore scale the
  proportional residual SD per subject (`propSdi <- propSd * exp(etapropSd)`)
  rather than treating it as a single population constant.
- **ELF residual error is fixed, not estimated.** Sect. 2.2: with one ELF
  observation per individual the ELF residual could not be identified and was
  fixed to the plasma median. Table 2 confirms `RESM ELF` carries CV 0%.
  Sect. 4 notes the corollary the authors themselves flag: the ELF
  between-subject variance estimates are *conditional* on that assumed residual,
  and would shrink if a larger residual had been assumed.
- **Penetration ratios are relative to TOTAL concentrations in both matrices.**
  Sect. 4 is explicit about this and about why it is conservative: plasma free
  fraction is about 85% (ceftazidime) and 92% (avibactam), ELF protein content is
  much lower than plasma, so the free fraction in ELF is likely higher than in
  plasma. Neither model file applies a protein-binding correction, and `Celf`
  should not be read as a free concentration without one. The PK/PD targets
  plotted in Figure 3 are *plasma-derived* (mouse data, references [10-12]) and
  the paper states they likely overestimate what is required in ELF.
- **`Celf` is an algebraic observable, not a compartment.** This follows the
  paper's model selection (Sect. 3.2, Sect. 3.3), not a simplification: the
  effect-site equilibration model was fitted, gave no OFV improvement, and
  implied ELF half-lives of 13 min (ceftazidime) and 8 min (avibactam). The
  `KELF` estimates behind those half-lives are quoted in prose only and are not
  tabulated, so the rejected effect-site model is not packaged.
- **No covariates.** The paper screened none and reports no demographic table
  beyond "43 healthy male volunteers", so neither model file declares
  `covariateData` and the virtual cohort carries no covariate columns.
- **Dosing history is reconstructed, not published per subject.** Sect. 2.1 gives
  nine 2 h infusions every 8 h; the vignette's event table reproduces exactly
  that. The paper's own simulations (Sect. 3.5, Figs. 2-3) used 1000 subjects on
  the 2000-500 mg regimen; this vignette uses 150 per arm, which is sufficient
  for a median comparison and stays under the 200/arm cap.
- **Literature `Cmax` comparison is external to this paper.** The 70 mg/L and
  12 mg/L values in the NCA comparison table come from reference [26] as quoted
  by Dimelow 2018, not from an NCA the authors performed. See the discussion of
  the starred row above.
- **`useLinCmt = FALSE` is mandatory** for every `rxSolve()` call against these
  models. The default ODE-to-`linCmt()` conversion drops `peripheral2` from a
  `k13`/`k31` parameterisation and mis-maps the two endpoints onto compartment
  slots.
