---
title: "Amodiaquine seasonal malaria chemoprevention (Ding 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Amodiaquine seasonal malaria chemoprevention (Ding 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Ding J, Coldiron ME, Assao B, Guindo O, Blessborn D, Winterberg M,
  Grais RF, Koscalova A, Langendorf C, Tarning J (2020). Adherence and
  population pharmacokinetic properties of amodiaquine when used for seasonal
  malaria chemoprevention in African children. *Clinical Pharmacology &
  Therapeutics* 107(5):1179-1188.
* Article: <https://doi.org/10.1002/cpt.1707> (open access, PMC7232861).
* The final NONMEM control stream is printed at the end of the paper's
  Supplementary Material S1 ("NONMEM code for final population PK model"). It
  is the primary source for the model structure and for every `ini()` value;
  the values agree with Table 1 of the main paper.

```{r load-model}
mod <- rxode2::rxode2(readModelDb("Ding_2020_amodiaquine")())
mod_typ <- rxode2::zeroRe(mod)
```

# Population

The PK cohort was a convenience sample of 136 children aged 3-59 months
(78/136, 57.4%, male) from households near the Epicentre study centre in
Magaria District, Niger. They received a directly observed course of seasonal
malaria chemoprevention (SMC) in November 2016, about 4.5 weeks after the
fourth monthly SMC round: a single dose of sulfadoxine-pyrimethamine with the
first of three once-daily amodiaquine doses (57.5 mg for ages 3-11 months,
153 mg for ages 12-59 months). Each child gave capillary blood (dried on
filter paper) in 3 of 6 pre-defined windows between 0 h and 35 days after the
first dose, 404 samples in total. Body weight was not measured; it was
predicted from age with a regression derived from 127,256 rural Nigerien
children (Supplementary Material S1, equations 2-4), and the predicted median
weight (10 kg) is the allometric reference. The four preceding SMC rounds were
assumed fully taken and included in the fit.

```{r population}
str(attr(mod, "metadata")$population)
```

# Source trace

| Item | Value | Source |
|---|---|---|
| Structure: first-order absorption, 2-cmt AQ, 3-cmt DEAQ, 100% molar conversion | ADVAN5, 6 compartments | S1 `$MODEL`/`$PK`; S1 Methods "Population PK analysis"; Figure 1 |
| `lka` | 2.85 1/h | Table 1; S1 THETA(3) |
| `lcl` | 101 L/h | Table 1; S1 THETA(1) |
| `lvc` | 314 L | Table 1; S1 THETA(2) |
| `lq` | 119 L/h | Table 1; S1 THETA(5) |
| `lvp` | 1,820 L | Table 1; S1 THETA(6) |
| `lfdepot` | 1 (held constant) | Table 1; S1 THETA(4) |
| `lcl_deaq` | 2.33 L/h | Table 1; S1 THETA(7) |
| `lvc_deaq` | 49.1 L | Table 1; S1 THETA(8) |
| `lq_deaq` | 2.31 L/h | Table 1 (S1 THETA(9) prints 2.3) |
| `lvp_deaq` | 363 L | Table 1; S1 THETA(10) |
| `lq2_deaq` | 4.34 L/h | Table 1; S1 THETA(11) |
| `lvp2_deaq` | 98.1 L | Table 1; S1 THETA(12) |
| `lpna50` | 4.66 months | Table 1; S1 THETA(13) |
| `lpna50_deaq` | 2.42 months | Table 1; S1 THETA(14) |
| `e_wt_cl`, `e_wt_vc` | 0.75, 1 | S1 equations 5-6; `$PK` `(WT/10)**0.75`, `(WT/10)` |
| Maturation `PNA / (pna50 + PNA)` | on CL AQ and CL DEAQ | S1 equation 7; Table 1 footnote |
| `etalcl`, `etalvc`, `etalka`, `etalfdepot` | 0.0493, 0.646, 3.0, 0.141 | S1 `$OMEGA` 1-4; Table 1 CV 22.2, 80.4, 173, 37.5% |
| `etalcl_deaq`, `etalvp_deaq` | 0.0232, 0.466 | S1 `$OMEGA` 7 and 10; Table 1 CV 15.2, 68.3% |
| `expSd`, `expSd_deaq` | 0.829, 0.204 | Table 1 sigma; S1 `$SIGMA` 0.688, 0.0417 (variances) |
| Output units nmol/L | `S2 = V2/1000` with AMT in umol | S1 `$INPUT`, `$PK` |
| Weight from age | WT = 5.46 + 0.162 * age (months) | S1 equation 4 |

# Virtual cohort

The paper gives the age range (3-59 months) and median (36 months) of the
PK cohort but not its distribution, so ages are drawn uniformly over whole
months 3-59. Weight is then predicted from age with the paper's own pooled
regression (Supplementary Material S1, equation 4), exactly as in the
analysis, and the dose follows the age band.

```{r cohort}
set.seed(2020)
rxode2::rxSetSeed(2020)
n_sub <- 200
cohort <- data.frame(id = seq_len(n_sub), PNA = sample(3:59, n_sub, replace = TRUE)) |>
  mutate(
    WT = 5.46 + 0.162 * PNA,
    dose = ifelse(PNA < 12, 57.5, 153),
    treatment = ifelse(PNA < 12, "57.5 mg (3-11 months)", "153 mg (12-59 months)")
  )
count(cohort, treatment)

obs_times <- sort(unique(c(seq(0, 72, by = 1), seq(78, 168, by = 6), seq(192, 24 * 84, by = 24))))
make_events <- function(subj, dose_times = c(0, 24, 48)) {
  doses <- tidyr::crossing(subj, time = dose_times) |>
    mutate(amt = dose, evid = 1L, cmt = "depot", dvid = NA_integer_)
  obs <- tidyr::crossing(subj, time = obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = NA_character_, dvid = 1L)
  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}
events <- make_events(cohort)
```

# Simulation

Observation rows carry `dvid = 1` with no `cmt`; rxode2 returns both
observables (`Cc` for amodiaquine, `Cc_deaq` for desethylamodiaquine) as
columns on every row.

```{r simulate}
sim <- rxode2::rxSolve(mod, events, keep = c("PNA", "WT", "treatment"),
                       returnType = "data.frame")
stopifnot(nrow(sim) > 0, all(is.finite(sim$Cc_deaq)))
```

# Replicate published figures

## Concentration-time profiles (cf. Figure 2)

Figure 2 of the paper is a VPC against the observed data, which are not
available; the simulated 5th, 50th and 95th percentiles for the 12-59 month
dose band (153 mg daily) are shown on the same log axes. Amodiaquine is shown
only to 300 h, the window the paper modelled for amodiaquine.

```{r fig-vpc, fig.cap = "Simulated 5th, 50th and 95th percentiles (153 mg daily for 3 days, ages 12-59 months); compare with Figure 2 of Ding 2020."}
pi_df <- sim |>
  filter(treatment == "153 mg (12-59 months)", time > 0) |>
  select(id, time, Amodiaquine = Cc, Desethylamodiaquine = Cc_deaq) |>
  pivot_longer(c(Amodiaquine, Desethylamodiaquine), names_to = "analyte", values_to = "conc") |>
  filter(analyte == "Desethylamodiaquine" | time <= 300) |>
  group_by(analyte, time) |>
  summarise(p05 = quantile(conc, 0.05), p50 = median(conc), p95 = quantile(conc, 0.95),
            .groups = "drop")
ggplot(pi_df, aes(time / 24, p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.25) +
  geom_line() +
  scale_y_log10() +
  facet_wrap(~analyte, scales = "free") +
  labs(x = "Time after first dose (days)", y = "Concentration (nmol/L)")
```

## Exposure versus age (cf. Figure S1a)

Figure S1a shows total desethylamodiaquine exposure after 10 mg/kg/day for 3
days by age, with the equation-4 weight assigned to each age. For a typical
child with complete conversion, total DEAQ exposure is exactly the molar dose
divided by `CL_DEAQ`, so the typical-value curve can be drawn in closed form
from the model's own clearance.

```{r fig-s1a, fig.cap = "Typical desethylamodiaquine AUC0-inf after 10 mg/kg/day for 3 days by age; compare with Figure S1a of Ding 2020."}
ages <- data.frame(id = 1:57, PNA = 3:59) |> mutate(WT = 5.46 + 0.162 * PNA, dose = 10 * WT)
ev_age <- ages |>
  tidyr::crossing(time = c(0, 24, 48)) |>
  mutate(amt = dose, evid = 1L, cmt = "depot", dvid = NA_integer_) |>
  bind_rows(mutate(ages, time = 72, amt = NA_real_, evid = 0L, cmt = NA_character_, dvid = 1L)) |>
  arrange(id, time, desc(evid))
typ_age <- rxode2::rxSolve(mod_typ, ev_age, keep = c("PNA", "WT"), returnType = "data.frame") |>
  filter(time == 72) |>
  mutate(auc_deaq = 3 * 10 * WT / 355.85 * 1e6 / cl_deaq) # mmol * 1e6 / (L/h) = h*nmol/L
ggplot(typ_age, aes(PNA, auc_deaq / 1000)) +
  geom_line() +
  labs(x = "Age (months)", y = "DEAQ AUC0-inf (h x umol/L)")
```

Exposure is highest in the youngest children because maturation lowers
clearance faster than weight-based dosing lowers the dose, the pattern the
paper describes.

# PKNCA validation

NCA is run on the individual predictions over the whole 3-day course (a
single interval from the first dose to infinity), which is how the paper's
Table 1 secondary parameters were derived (from the empirical Bayes
estimates, without residual error).

```{r pknca}
nca_in <- sim |>
  select(id, time, treatment, Amodiaquine = Cc, Desethylamodiaquine = Cc_deaq) |>
  pivot_longer(c(Amodiaquine, Desethylamodiaquine), names_to = "analyte", values_to = "conc") |>
  filter(!is.na(conc)) |>
  filter(analyte == "Desethylamodiaquine" | time <= 300)
conc_obj <- PKNCA::PKNCAconc(nca_in, conc ~ time | treatment + analyte + id)
intervals <- data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE,
                        half.life = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, intervals = intervals))
nca_tab <- as.data.frame(nca_res)
summary(nca_res)
```

## Closed-form mass-balance gate

With complete molar conversion, each subject's desethylamodiaquine
AUC0-inf equals `F * total molar dose / CL_DEAQ` exactly. The NCA value
differs only by trapezoidal and extrapolation error.

```{r mass-balance}
par_ind <- sim |>
  group_by(id) |>
  slice(1) |>
  ungroup() |>
  select(id, cl_deaq, fdepot) |>
  left_join(cohort, by = "id")
chk <- nca_tab |>
  filter(analyte == "Desethylamodiaquine", PPTESTCD == "aucinf.obs") |>
  mutate(id = as.integer(as.character(id))) |>
  left_join(par_ind, by = "id") |>
  mutate(auc_exact = fdepot * 3 * dose / 355.85 * 1e6 / cl_deaq,
         pct_diff = 100 * (PPORRES - auc_exact) / auc_exact)
summary(chk$pct_diff)
stopifnot(
  abs(median(chk$pct_diff)) < 2,
  quantile(abs(chk$pct_diff), 0.9) < 5
)
```

## Comparison against published secondary parameters

Table 1 reports medians over the empirical Bayes estimates of the 136
children (AUC in h x umol/L, converted here to h x nmol/L). The terminal
half-lives are compared as NCA values.

```{r nca-compare}
ref <- data.frame(
  analyte = c("Amodiaquine", "Desethylamodiaquine"),
  cmax = c(835, 3272),
  aucinf.obs = c(14.76e3, 576e3),
  half.life = c(27.1, 11.0 * 24)
)
sim_cmp <- nca_tab |>
  filter(PPTESTCD %in% c("cmax", "aucinf.obs", "half.life")) |>
  select(analyte, PPTESTCD, PPORRES)
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_cmp,
  reference = ref,
  by = "analyte",
  units = c(cmax = "nmol/L", aucinf.obs = "h*nmol/L", half.life = "h"),
  tolerance_pct = 20
)
knitr::kable(cmp, digits = 1, caption = paste(
  "Simulated cohort median versus Ding 2020 Table 1 (median of empirical Bayes estimates).",
  attr(cmp, "footnote")
))
```

Desethylamodiaquine Cmax, both AUCs and both half-lives agree with Table 1
within 20%. Amodiaquine Cmax is flagged: the simulated cohort median is
about a third below the published median. That published value is a median
of empirical Bayes estimates, and the absorption rate constant carries a
173% CV with 42.8% shrinkage (Table 1), so the individual estimates are
pulled toward the typical absorption rate while a stochastic simulation
draws the full spread of slow absorbers, which blunts the amodiaquine peak.
Desethylamodiaquine, formed slowly, is insensitive to this. The typical
child (36 months, 153 mg daily, random effects set to zero) reproduces the
published amodiaquine Cmax closely:

```{r typical-cmax}
ev_typ <- make_events(data.frame(id = 1, PNA = 36, WT = 5.46 + 0.162 * 36, dose = 153,
                                 treatment = "typical"))
sim_typ <- rxode2::rxSolve(mod_typ, ev_typ, returnType = "data.frame")
c(Cmax_AQ = max(sim_typ$Cc), Cmax_DEAQ = max(sim_typ$Cc_deaq))
```

# Adherence methodology (Table 3 and Results)

The paper simulated a typical child (36 months, 11.5 kg, 153 mg daily)
under the 4 (first-dose DOT) and 8 (non-DOT) dosing scenarios of Table 3,
took the desethylamodiaquine concentration at a given percentile of the
fully adherent scenario as the cut-off, and reported the sensitivity (the
fraction of simulated non-adherent children falling below it). The cut-off
percentiles are the Youden-optimal values the paper reports. Residual error
is added to the individual predictions because the method is applied to
measured concentrations; scenario 8 (no dose) is assigned zero. The paper
used 2,000 children per scenario; 200 are used here. With only 200 fully
adherent children the cut-off quantile itself is noisy (it moves the
sensitivity by several percentage points), so the scenarios share common
random numbers: the same 200 children (the same random effects, reseeded
before each scenario's solve, and the same residual draws) are simulated
under every dosing pattern, and only the doses differ.

```{r adherence}
scen <- list(S1 = c(1, 1, 1), S2 = c(1, 1, 0), S3 = c(1, 0, 1), S4 = c(1, 0, 0),
             S5 = c(0, 1, 1), S6 = c(0, 1, 0), S7 = c(0, 0, 1), S8 = c(0, 0, 0))
n_arm <- 200
days <- c(3, 7, 14, 21, 28)
eps_adh <- rnorm(n_arm * length(days), 0, 0.204)
solve_scenario <- function(dose_t) {
  obs <- expand.grid(id = seq_len(n_arm), time = 24 * days) |>
    mutate(amt = NA_real_, evid = 0L, cmt = NA_character_, dvid = 1L)
  dos <- expand.grid(id = seq_len(n_arm), time = dose_t) |>
    mutate(amt = 153, evid = 1L, cmt = "depot", dvid = NA_integer_)
  ev <- bind_rows(dos, obs) |>
    mutate(PNA = 36, WT = 11.5) |>
    arrange(id, time, desc(evid))
  rxode2::rxSetSeed(2021)
  s <- rxode2::rxSolve(mod, ev, returnType = "data.frame") |> arrange(id, time)
  s$obs <- s$Cc_deaq * exp(eps_adh)
  s
}
sim_adh <- bind_rows(lapply(names(scen)[1:7], function(k) {
  mutate(solve_scenario(c(0, 24, 48)[scen[[k]] == 1]), scenario = k)
}))
# Scenario 8 (no dose taken) has no drug: every concentration is zero.
sim_adh <- bind_rows(sim_adh, filter(sim_adh, scenario == "S1") |>
  mutate(scenario = "S8", Cc_deaq = 0, obs = 0))

sensitivity <- function(pct, day, nonadherent) {
  x <- sim_adh[sim_adh$time == 24 * day, ]
  cutoff <- quantile(x$obs[x$scenario == "S1"], pct / 100)
  100 * mean(x$obs[x$scenario %in% nonadherent] < cutoff)
}
dot <- c("S2", "S3", "S4")
nondot <- c("S2", "S3", "S4", "S5", "S6", "S7", "S8")
adh <- data.frame(
  regimen = rep(c("First-dose DOT", "Non-DOT"), each = 4),
  day = rep(c(3, 7, 14, 28), 2),
  cutoff_pct = c(20, 20, 25, 30, 20, 20, 20, 25),
  paper_sens = c(71, 65, 66, 68, 77, 73, 71, 72)
)
adh$sim_sens <- mapply(function(p, d, r) sensitivity(p, d, if (r == "Non-DOT") nondot else dot),
                       adh$cutoff_pct, adh$day, adh$regimen)
adh |>
  dplyr::rename("Regimen" = regimen, "Day" = day, "Cut-off percentile" = cutoff_pct,
                "Paper sensitivity (%)" = paper_sens, "Simulated sensitivity (%)" = sim_sens) |>
  knitr::kable(digits = 0, caption = "Sensitivity at the optimal cut-off percentiles (Ding 2020 Results).")
stopifnot(
  abs(median(adh$sim_sens - adh$paper_sens)) < 5,
  all(abs(adh$sim_sens - adh$paper_sens) < 10)
)
```

At the conservative 5th-percentile cut-off the paper reports sensitivities
of 22-45% (DOT) and 35-55% (non-DOT):

```{r adherence-p5}
p5 <- sapply(days, function(d) c(DOT = sensitivity(5, d, dot), nonDOT = sensitivity(5, d, nondot)))
colnames(p5) <- paste("Day", days)
knitr::kable(p5, digits = 0, caption = "Simulated sensitivity (%) at the 5th-percentile cut-off.")
```

The simulated ranges span the published ones, running a few points
wider at both ends: sensitivity is highest shortly after dosing and falls
as the desethylamodiaquine profiles of the scenarios converge.

The paper's dose-timing sensitivity analysis compared the day-28
5th-percentile cut-off after all three doses on day 1 with full adherence
and found a relative difference below 23% (about 10 nmol/L):

```{r dose-timing}
sim_t <- solve_scenario(c(0, 0, 0))
p5_same_day <- quantile(sim_t$obs[sim_t$time == 672], 0.05)
p5_full <- quantile(sim_adh$obs[sim_adh$scenario == "S1" & sim_adh$time == 672], 0.05)
c(full_adherence = unname(p5_full), all_on_day1 = unname(p5_same_day),
  relative_diff_pct = unname(100 * (p5_same_day - p5_full) / p5_full))
```

Taking all three doses on day 1 lowers the day-28 cut-off by a relative
difference (printed above) inside the paper's bound of 23%.

# Assumptions and deviations

* **Typographical slips in the printed control stream.** The Supplementary
  Material S1 `$PK` block has three evident slips, corrected here: (i) the
  maturation factors are defined as `MF_AQ` / `MF_DEAQ` but applied as
  `MF1` / `MF2`; (ii) `F1 = TVF1 * EXP(ETA4)` is read as `EXP(ETA(4))`;
  (iii) `Q1 = TVQ1 * EXP(ETA(3))` would give the amodiaquine
  inter-compartmental clearance the absorption-rate random effect. It is read
  as `EXP(ETA(5))`, which is held at zero, because every other parameter uses
  its own sequential eta index, `OMEGA(5)` is the zero-valued "AQ IIV
  inter-compartment clearance" entry, and Table 1 reports no IIV on Q/F AQ.
  So Q/F AQ carries no IIV in this model.
* **Omega values.** The S1 `$OMEGA` and `$SIGMA` blocks are headed "Initial
  estimates" but are the final estimates: the square root of each value
  reproduces the Table 1 "CV for IIV" and sigma entries to the printed
  precision. (Table 1's footnote gives the CV as `sqrt(exp(omega) - 1)`, but
  its numbers are `sqrt(omega)`; the variances themselves are used here.)
* **Q1/F DEAQ.** Table 1 prints 2.31 L/h and the control stream 2.3; the
  Table 1 value is used.
* **Molar conversion.** The paper fitted molar amounts (dose in umol). The
  model takes doses in mg of amodiaquine base and applies the base molecular
  weights (355.85 and 327.81 g/mol, as in the library's other amodiaquine
  models) to form the metabolite and report nmol/L; this is numerically
  identical to the paper's molar system.
* **Weight is a predicted covariate.** The paper never measured weight; users
  reproducing the analysis should derive `WT` from age with equation 4
  (`WT = 5.46 + 0.162 * PNA`), as done in this article. Measured weights can
  be supplied instead, but that is an extrapolation of the model.
* **Specimen.** Concentrations are capillary whole blood dried on filter
  paper, not plasma.
* **Age distribution.** Only the range and median of the PK cohort are
  reported; a uniform age distribution over 3-59 months is assumed for the
  virtual cohort, which puts its median age (and hence exposure) somewhat
  below the paper's cohort.
* **Adherence day definition.** "Day 3" etc. is taken as 72 h etc. after the
  first dose; the typical child of the adherence analysis uses the paper's
  stated 11.5 kg rather than the equation-4 weight at 36 months (11.3 kg).
* **Previous SMC rounds.** The fit assumed the four previous monthly rounds
  were taken; the simulations here start from a drug-free state, which omits
  a small residual desethylamodiaquine carry-over.
* **Not retained covariates.** Weight-for-age Z-score on bioavailability and
  sex were tested but not retained; they are documented in the model's
  `covariatesDataExcluded` metadata.
