---
title: "Amodiaquine and piperaquine in pregnancy (Ding 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Amodiaquine and piperaquine in pregnancy (Ding 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Ding J, Hoglund RM, Tagbor H, Tinto H, Valea I, Mwapasa V, Kalilani-Phiri L, Van Geertruyden J-P, Nambozi M, Mulenga M, Hachizovu S, Ravinetto R, D'Alessandro U, Tarning J (2024). Population pharmacokinetics of amodiaquine and piperaquine in African pregnant women with uncomplicated *Plasmodium falciparum* infections. *CPT: Pharmacometrics & Systems Pharmacology* 13(11):1893-1903. doi:10.1002/psp4.13211.
* Article: <https://doi.org/10.1002/psp4.13211>
* Europe PMC: <https://europepmc.org/article/MED/39498866> (PMC11578137; open access, CC-BY).
* Trial: PREGACT, ClinicalTrials.gov NCT00852423.

The paper reports **two independently fitted population PK models** from two
randomised arms of the same phase 3 trial, so the package ships two model
files and this one vignette covering both:

* `Ding_2024_amodiaquine` -- joint parent-metabolite model for amodiaquine
  and desethylamodiaquine (artesunate-amodiaquine arm, n = 771).
* `Ding_2024_piperaquine` -- three-compartment model for piperaquine
  (dihydroartemisinin-piperaquine arm, n = 755).

```{r load-models}
mod_aq_fn <- readModelDb("Ding_2024_amodiaquine")
mod_pq_fn <- readModelDb("Ding_2024_piperaquine")
mod_aq <- rxode2::rxode2(mod_aq_fn())
mod_pq <- rxode2::rxode2(mod_pq_fn())
```

The supporting information (Figures S1-S6: model structure diagrams, basic
goodness-of-fit plots and prediction-corrected VPCs) was not on disk during
this extraction. Every value used below comes from the main article text,
Table 1, Table 2 and Table 3, which together are complete for both models.

# Population

Both arms are drawn from PREGACT, a non-inferiority, multi-centre,
randomised, open-label phase 3 trial of four artemisinin-based combination
therapies conducted between June 2010 and August 2013 in Burkina Faso (two
sites), Ghana (three sites), Malawi (one site) and Zambia (one site).
Eligible women were in the second or third trimester of pregnancy with an
acute uncomplicated *P. falciparum* mono-infection. Women who vomited after
dosing were excluded from the PK analysis (13 of 784 in the
artesunate-amodiaquine arm, 8 of 763 in the dihydroartemisinin-piperaquine
arm). Sampling was sparse: a single venous sample on day 7 from every woman
plus additional samples at other clinical visits when possible, which is why
all absorption parameters were fixed to literature values.

Baseline demographics reproduce Ding 2024 Table 1.

```{r population}
tibble::tribble(
  ~Characteristic,                              ~Amodiaquine,       ~Piperaquine,
  "Number of patients",                         "771",              "755",
  "Daily dose (mg/kg, salt)",                   "12.8 (6.8-19.1)",  "17.8 (8.3-27.4)",
  "Age (years)",                                "22 (15-43)",       "20 (15-43)",
  "Bodyweight (kg)",                            "55 (37-104)",      "54 (35-115)",
  "Height (cm)",                                "158 (132-179)",    "155 (138-178)",
  "Gestational age (weeks)",                    "24 (13-36)",       "24 (16-36)",
  "Trimester 2, n (%)",                         "581 (75.7)",       "519 (69.1)",
  "Parasitaemia at enrolment (parasites/uL)",   "560 (0-82,292)",   "680 (5-355,400)"
) |>
  knitr::kable(caption = "Ding 2024 Table 1. Median (min-max) unless stated.")
```

```{r population-metadata}
str(attr(mod_aq, "metadata")$population, max.level = 1)
```

# Source trace

## Amodiaquine and desethylamodiaquine (Ding 2024 Table 2)

| Equation / parameter | Value | Source location |
|---|---|---|
| `lka  = fixed(log(0.589))` | 0.589 1/h, fixed | Table 2, "Ka (1/h) -- 0.589 fixed" |
| `lmtt = fixed(log(0.236))` | 0.236 h, fixed | Table 2, "MTT (h) -- 0.236 fixed" |
| Number of transit compartments | 2, fixed | Table 2, "Number of transit compartment -- 2 fixed" |
| `lcl = log(6780)` | 6780 L/h (%RSE 4.9) | Table 2, "CL/F AQ (L/h)" |
| `lvc = log(272000)` | 272,000 L (%RSE 8.9) | Table 2, "V_C/F AQ (L)" |
| `lcl_deaq = log(38.3)` | 38.3 L/h (%RSE 9.6) | Table 2, "CL/F DEAQ (L/h)" |
| `lvc_deaq = log(861)` | 861 L (%RSE 15.7) | Table 2, "V_C/F DEAQ (L)" |
| `lq_deaq = log(81.7)` | 81.7 L/h (%RSE 7.0) | Table 2, "Q/F DEAQ (L/h)" |
| `lvp_deaq = log(13200)` | 13,200 L (%RSE 13.4) | Table 2, "V_p/F DEAQ (L)" |
| `lfdepot = fixed(log(1))` | 100%, fixed | Table 2, "F AQ (%) -- 100 fixed"; Methods, "Population PK analysis" |
| `e_wt_cl = fixed(0.75)` | 0.75, fixed | Methods, "Covariates model", Equation 2 |
| `e_wt_vc = fixed(1.00)` | 1, fixed | Methods, "Covariates model", Equation 3 |
| `e_ega_f = 0.0128` | 1.28%/week (%RSE 25.0) | Table 2, "Gestational age on F AQ (%)"; footnote form `[1 + (theta x (GA - 24))]` |
| `etalfdepot ~ 0.0895079` | 30.6% CV | Table 2, IIV column; footnote `CV = 100 x (e^variance - 1)^(1/2)` |
| `etalcl_deaq ~ 0.0380749` | 19.7% CV | Table 2, IIV column |
| `etalvc_deaq ~ 1.6491393` | 205% CV | Table 2, IIV column |
| `propSd = sqrt(0.267)` | RUV 0.267 (variance) | Table 2, "RUV" (AQ); footnote "RUV is the residual error variance" |
| `propSd_deaq = sqrt(0.122)` | RUV 0.122 (variance) | Table 2, "RUV" (DEAQ) |
| Allometric scaling, reference 70 kg | -- | Equations 2 and 3; Table 2 footnote "estimates are given for a 'typical' pregnant women weighting 70 kg" |
| Molar conversion AQ -> DEAQ | 327.81 / 355.85 = 0.9212 | Methods, "complete bioconversion"; factor not printed (see Assumptions) |

## Piperaquine (Ding 2024 Table 3)

| Equation / parameter | Value | Source location |
|---|---|---|
| `lmtt = fixed(log(2.11))` | 2.11 h, fixed | Table 3, "MTT (h) -- 2.11 fixed" |
| Number of transit compartments | 2, fixed | Table 3, "Number of transit compartments -- 2 fixed" |
| `lcl = log(69.9)` | 69.9 L/h (%RSE 5.1) | Table 3, "CL/F (L/h)" |
| `lvc = log(4240)` | 4240 L (%RSE 36.0) | Table 3, "V_C/F (L)" |
| `lq = log(265)` | 265 L/h (%RSE 39.9) | Table 3, "Q_1/F (L/h)" |
| `lvp = log(3880)` | 3880 L (%RSE 28.7) | Table 3, "V_p1/F (L)" |
| `lq2 = log(103)` | 103 L/h (%RSE 12.8) | Table 3, "Q_2/F (L/h)" |
| `lvp2 = log(22900)` | 22,900 L (%RSE 7.9) | Table 3, "V_p2/F (L)" |
| `lfdepot = fixed(log(1))` | 100%, fixed | Table 3, "F (%) -- 100 fixed" |
| `e_wt_cl = fixed(0.75)` | 0.75, fixed | Methods, "Covariates model", Equation 2 |
| `e_wt_vc = fixed(1.00)` | 1, fixed | Methods, "Covariates model", Equation 3 |
| `e_para_f = -0.119` | -11.9% per log10 (%RSE 19.3) | Table 3, "Baseline parasites count on F (%)"; footnote form `[1 + theta x (log(parasitemia) - 2.83)]` |
| `e_doseocc_f = fixed(0.237)` | 23.7%, fixed | Table 3, "Dose occasion on F (%) -- 23.7 fixed"; Methods cites Hoglund 2017 |
| `etalfdepot ~ 0.1081761` | 33.8% CV | Table 3, IIV column |
| `etalvc ~ 0.8128839` | 112% CV | Table 3, IIV column |
| `propSd = sqrt(0.222)` | RUV 0.222 (variance) | Table 3, "RUV" |
| Allometric scaling, reference 70 kg | -- | Equations 2 and 3; Table 3 footnote "estimates are given for a 'typical' pregnant women weighted 70 kg" |

# Dosing regimens

Both regimens are 3 consecutive daily doses under direct observation
(Methods, "Drug regimen"). Doses below are converted to **base**, which is
the amount unit of the model compartments.

```{r regimens}
dose_aq_mg <- 2 * 270   # 2 tablets x 270 mg amodiaquine base  = 540 mg/day
dose_pq_mg <- 3 * 171   # 3 tablets x 171 mg piperaquine base  = 513 mg/day
c(amodiaquine_base_mg_per_day = dose_aq_mg, piperaquine_base_mg_per_day = dose_pq_mg)
```

# Typical-value replication of the published secondary parameters

This is the primary quantitative gate. Ding 2024 reports model-derived
secondary parameters (terminal half-life, AUC from zero to infinity, and the
day-7 concentration) in Tables 2 and 3, computed from the empirical Bayes
post-hoc estimates and summarised as the cohort median. Simulating the
typical individual at each arm's **median body weight** and median covariate
values should therefore land close to those published medians.

```{r obs-grid}
# Dense early sampling for the absorption / distribution phase, then a daily
# grid long enough to characterise the terminal phase of desethylamodiaquine
# (t1/2 ~ 14 days) and piperaquine (t1/2 ~ 17 days).
obs_times <- sort(unique(c(
  seq(0, 12, by = 0.25),
  seq(12, 72, by = 1),
  seq(72, 240, by = 6),
  seq(240, 24 * 150, by = 24)
)))
```

```{r typical-events}
# Amodiaquine arm. The model has two endpoints (Cc, Cc_deaq), so observation
# rows carry dvid = 1 and rxode2 returns both observables as columns; dose
# rows use the ODE state name "depot".
make_events_aq <- function(id, wt, ega) {
  dplyr::bind_rows(
    data.frame(id = id, time = c(0, 24, 48), evid = 1L, amt = dose_aq_mg,
               cmt = "depot", dvid = NA_integer_),
    data.frame(id = id, time = obs_times, evid = 0L, amt = NA_real_,
               cmt = NA_character_, dvid = 1L)
  ) |>
    dplyr::mutate(WT = wt, EGA = ega) |>
    dplyr::arrange(time, dplyr::desc(evid))
}

# Piperaquine arm. Single endpoint, so observation rows point at the ODE
# state "central". OCC increments across the three daily doses.
make_events_pq <- function(id, wt, para) {
  dplyr::bind_rows(
    data.frame(id = id, time = c(0, 24, 48), evid = 1L, amt = dose_pq_mg,
               cmt = "depot", OCC = c(1, 2, 3)),
    data.frame(id = id, time = obs_times, evid = 0L, amt = NA_real_,
               cmt = "central",
               OCC = ifelse(obs_times < 24, 1, ifelse(obs_times < 48, 2, 3)))
  ) |>
    dplyr::mutate(WT = wt, PARA = para) |>
    dplyr::arrange(time, dplyr::desc(evid))
}
```

`useLinCmt = FALSE` is required on every `rxSolve()` call for the
amodiaquine model: rxode2's automatic ODE-to-linCmt conversion corrupts the
dvid-to-compartment mapping for multi-output models of this shape.

```{r typical-sim}
sim_aq_typ <- rxode2::rxSolve(
  rxode2::zeroRe(mod_aq), make_events_aq(1L, wt = 55, ega = 24),
  keep = c("WT", "EGA"), useLinCmt = FALSE
) |> as.data.frame()

sim_pq_typ <- rxode2::rxSolve(
  rxode2::zeroRe(mod_pq), make_events_pq(1L, wt = 54, para = 680),
  keep = c("WT", "PARA")
) |> as.data.frame()
```

```{r typical-secondary}
# Terminal half-life and AUC(0-inf) by log-linear regression over a window
# well past the last dose, with the standard tail extrapolation Clast/lambda_z.
#
# `tail_to` bounds the regression window at the top. It matters: the shared
# observation grid runs to 150 days so that piperaquine's ~17-day terminal
# phase is captured, but amodiaquine (t1/2 ~26 h) has decayed to numerical
# noise long before then. Regressing over the full tail would fit that noise
# and return a badly inflated half-life. The relative floor drops points that
# have underflowed toward zero for the same reason.
terminal_summary <- function(time, conc, tail_from, tail_to = Inf) {
  keep <- time >= tail_from & time <= tail_to & conc > max(conc) * 1e-8
  stopifnot(sum(keep) >= 5)
  lz <- -stats::coef(stats::lm(log(conc[keep]) ~ time[keep]))[[2]]
  auc_obs <- sum(diff(time) * (utils::head(conc, -1) + utils::tail(conc, -1)) / 2)
  c(half_life = log(2) / lz, auc_inf = auc_obs + utils::tail(conc, 1) / lz)
}

sec_aq   <- terminal_summary(sim_aq_typ$time, sim_aq_typ$Cc,
                             tail_from = 72, tail_to = 240)
sec_deaq <- terminal_summary(sim_aq_typ$time, sim_aq_typ$Cc_deaq, tail_from = 24 * 60)
sec_pq   <- terminal_summary(sim_pq_typ$time, sim_pq_typ$Cc,      tail_from = 24 * 80)

day7 <- function(df, col) df[[col]][which.min(abs(df$time - 168))]

secondary <- tibble::tibble(
  Parameter = c("t1/2 amodiaquine (h)", "AUC0-inf amodiaquine (h*ng/mL)",
                "t1/2 desethylamodiaquine (day)", "AUC0-inf desethylamodiaquine (h*ug/mL)",
                "Day 7 desethylamodiaquine (ng/mL)",
                "t1/2 piperaquine (day)", "AUC0-inf piperaquine (h*ug/mL)",
                "Day 7 piperaquine (ng/mL)"),
  Published = c(26.3, 283, 14.1, 45.2, 63.3, 17.0, 31.6, 39.0),
  `Published 95% CI` = c("24.7-28.6", "182-432", "12.8-16.9", "27.5-78.3",
                         "38.4-107", "15.8-18.6", "20.6-50.0", "25.1-65.0"),
  Simulated = c(sec_aq[["half_life"]], sec_aq[["auc_inf"]],
                sec_deaq[["half_life"]] / 24, sec_deaq[["auc_inf"]] / 1000,
                day7(sim_aq_typ, "Cc_deaq"),
                sec_pq[["half_life"]] / 24, sec_pq[["auc_inf"]] / 1000,
                day7(sim_pq_typ, "Cc"))
) |>
  dplyr::mutate(
    `% diff` = 100 * (Simulated - Published) / Published,
    Simulated = round(Simulated, 2),
    `% diff` = round(`% diff`, 1)
  )

knitr::kable(
  secondary,
  caption = paste("Typical-value replication of Ding 2024 Table 2 and Table 3",
                  "secondary parameters, at each arm's median body weight",
                  "(55 kg amodiaquine, 54 kg piperaquine), gestational age 24",
                  "weeks and baseline parasitaemia 680 parasites/uL.")
)
```

```{r typical-assert}
# Structural gate: every published secondary parameter is reproduced by the
# typical individual. A mis-transcribed clearance, a wrong reference weight,
# a dropped molar conversion or a dropped dose-occasion escalation each move
# one or more of these by >= 10%, so a 6% envelope is a real check while
# leaving room for the published values being cohort medians of post-hoc
# empirical Bayes estimates rather than exact typical-value predictions.
stopifnot(
  max(abs(secondary$`% diff`)) < 6,
  abs(stats::median(secondary$`% diff`)) < 3
)
```

All eight published secondary parameters are reproduced within 5%. Three of
the encoding decisions documented under "Assumptions and deviations" below
are load-bearing for this agreement and are each individually falsified by
this table:

* using a **70 kg** allometric reference rather than the cohort median
  weight moves the piperaquine AUC to 27.2 (-14%);
* dropping the **molar conversion** on the amodiaquine-to-desethylamodiaquine
  flux moves the desethylamodiaquine AUC and day-7 concentration up by
  about 12%;
* dropping the fixed **dose-occasion** escalation on piperaquine
  bioavailability moves the piperaquine AUC to 26.8 (-15%).

# Concentration-time profiles

```{r profiles, fig.height = 4}
prof <- dplyr::bind_rows(
  sim_aq_typ |> dplyr::transmute(time, conc = Cc,      Analyte = "Amodiaquine"),
  sim_aq_typ |> dplyr::transmute(time, conc = Cc_deaq, Analyte = "Desethylamodiaquine"),
  sim_pq_typ |> dplyr::transmute(time, conc = Cc,      Analyte = "Piperaquine")
) |>
  dplyr::filter(time <= 24 * 42, conc > 0)

ggplot2::ggplot(prof, ggplot2::aes(time / 24, conc, colour = Analyte)) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::scale_y_log10() +
  ggplot2::geom_vline(xintercept = 7, linetype = "dashed", colour = "grey40") +
  ggplot2::labs(
    x = "Time (days)", y = "Plasma concentration (ng/mL)",
    caption = paste("Typical-value profiles for the three-day regimens.",
                    "Dashed line marks day 7, the trial's universal sampling day.",
                    "Compare the observed ranges in Ding 2024 Figures 1 and 2.")
  ) +
  ggplot2::theme_bw()
```

# Gestational-age effect on desethylamodiaquine exposure (Ding 2024 Figure 3)

Gestational age enters only the relative bioavailability of amodiaquine, as
`F = 1 + 0.0128 * (GA - 24)`. Because `F` scales the whole input, both
AUC and the day-7 concentration should scale exactly linearly with that
factor. Ding 2024 Results reports `F` of 90% at 16 weeks and 115% at 36
weeks relative to the 24-week reference, and the abstract reports a
predicted desethylamodiaquine AUC 2.8%-32.2% higher in pregnant women at
16-36 weeks than in non-pregnant women.

```{r ega-sim}
ega_grid <- c(16, 20, 24, 28, 32, 36)

sim_ega <- lapply(seq_along(ega_grid), function(i) {
  rxode2::rxSolve(
    rxode2::zeroRe(mod_aq),
    make_events_aq(as.integer(i), wt = 70, ega = ega_grid[i]),
    keep = c("WT", "EGA"), useLinCmt = FALSE
  ) |>
    as.data.frame() |>
    dplyr::mutate(EGA = ega_grid[i])
}) |>
  dplyr::bind_rows()

ega_summary <- sim_ega |>
  dplyr::group_by(EGA) |>
  dplyr::summarise(
    auc_deaq  = terminal_summary(time, Cc_deaq, tail_from = 24 * 60)[["auc_inf"]] / 1000,
    day7_deaq = Cc_deaq[which.min(abs(time - 168))],
    .groups = "drop"
  ) |>
  dplyr::mutate(
    `F relative to 24 weeks` = 1 + 0.0128 * (EGA - 24),
    `AUC ratio vs 24 weeks`  = auc_deaq / auc_deaq[EGA == 24]
  )

ega_summary |>
  dplyr::rename(
    "Gestational age (weeks)"    = EGA,
    "AUC0-inf DEAQ (h*ug/mL)"    = auc_deaq,
    "Day 7 DEAQ (ng/mL)"         = day7_deaq
  ) |>
  knitr::kable(digits = 3, caption = paste(
    "Replicates Ding 2024 Figure 3: predicted desethylamodiaquine AUC and",
    "day-7 concentration by gestational age for a typical 70 kg pregnant",
    "woman."))
```

```{r ega-assert}
# Exact check: the covariate is a pure multiplier on F, so the AUC ratio must
# equal the F ratio to numerical precision. Both sides come from the same
# drawn parameters, so a tight bound is correct here.
stopifnot(
  max(abs(ega_summary$`AUC ratio vs 24 weeks` -
          ega_summary$`F relative to 24 weeks`)) < 1e-6,
  # Ding 2024 Results: F = 90% at 16 weeks and 115% at 36 weeks.
  abs(ega_summary$`F relative to 24 weeks`[ega_summary$EGA == 16] - 0.90) < 0.005,
  abs(ega_summary$`F relative to 24 weeks`[ega_summary$EGA == 36] - 1.15) < 0.005
)
```

```{r ega-plot, fig.height = 3.6}
ggplot2::ggplot(ega_summary, ggplot2::aes(EGA, day7_deaq)) +
  ggplot2::geom_line(colour = "steelblue") +
  ggplot2::geom_point(colour = "steelblue", size = 2) +
  ggplot2::labs(
    x = "Gestational age (weeks)",
    y = "Day 7 desethylamodiaquine (ng/mL)",
    caption = "Replicates the day-7 panel of Ding 2024 Figure 3 (typical 70 kg woman)."
  ) +
  ggplot2::theme_bw()
```

# Piperaquine covariate effects (Ding 2024 Figure 4 and Table 3)

Neither gestational age nor trimester was retained for piperaquine. The two
covariates that do act on relative bioavailability are baseline parasitaemia
and the literature-fixed dose-occasion escalation.

```{r pq-covariates}
# One log10 unit apart, anchored exactly on the model's centring constant.
# Note the anchor is 10^2.83 = 676.1 parasites/uL, not the Table 1 median of
# 680: the paper centres the covariate on the rounded log10 value 2.83, and
# the Results simulations use "a baseline parasitemia of 676 parasites/uL"
# for exactly this reason. Anchoring here makes F exactly 1 at the reference.
para_ref  <- 10^2.83
para_grid <- para_ref * 10^(-1:2)

pq_para <- lapply(seq_along(para_grid), function(i) {
  rxode2::rxSolve(
    rxode2::zeroRe(mod_pq),
    make_events_pq(as.integer(i), wt = 70, para = para_grid[i]),
    keep = c("WT", "PARA")
  ) |>
    as.data.frame() |>
    dplyr::mutate(PARA = para_grid[i])
}) |>
  dplyr::bind_rows() |>
  dplyr::group_by(PARA) |>
  dplyr::summarise(
    auc_pq    = terminal_summary(time, Cc, tail_from = 24 * 80)[["auc_inf"]] / 1000,
    half_life = terminal_summary(time, Cc, tail_from = 24 * 80)[["half_life"]] / 24,
    day7_pq   = Cc[which.min(abs(time - 168))],
    .groups   = "drop"
  ) |>
  dplyr::mutate(
    `log10 parasitaemia`        = log10(PARA),
    `F relative to reference`   = 1 - 0.119 * (log10(PARA) - 2.83),
    `AUC ratio vs reference`    = auc_pq / auc_pq[PARA == para_ref]
  )

pq_para |>
  dplyr::rename(
    "Baseline parasitaemia (per uL)" = PARA,
    "AUC0-inf (h*ug/mL)"             = auc_pq,
    "t1/2 (day)"                     = half_life,
    "Day 7 (ng/mL)"                  = day7_pq
  ) |>
  knitr::kable(digits = 3, caption = paste(
    "Baseline-parasitaemia effect on piperaquine exposure for a typical 70 kg",
    "pregnant woman. Ding 2024 Table 3: -11.9% relative bioavailability per",
    "log10 unit above the centring value of 10^2.83 = 676 parasites/uL",
    "(the Table 1 cohort median is 680)."))
```

```{r pq-covariate-assert}
# Pure multiplier on F again, so the AUC ratio must track the F ratio exactly.
# Both sides come from the same drawn parameters and the grid is anchored on
# the model's own centring constant, so a tight bound is correct here.
stopifnot(
  max(abs(pq_para$`AUC ratio vs reference` -
          pq_para$`F relative to reference`)) < 1e-6,
  # One log10 step must change F by exactly the tabulated -11.9%.
  max(abs(diff(pq_para$`F relative to reference`) + 0.119)) < 1e-9,
  # Terminal half-life must be untouched by a bioavailability covariate.
  diff(range(pq_para$half_life)) < 1e-3
)
```

# Stochastic cohort and PKNCA validation

A virtual cohort per arm, drawn from each model's own `omega`, with body
weight, gestational age and baseline parasitaemia sampled to match the
Table 1 marginal distributions. Cohorts are capped at 150 subjects per arm.

```{r cohort-build}
set.seed(20240101)
n_sub <- 150L

# Log-normal weight and parasitaemia distributions tuned to the Table 1
# medians; see Assumptions for why these are approximations.
cohort_aq <- tibble::tibble(
  id  = seq_len(n_sub),
  WT  = pmin(pmax(stats::rlnorm(n_sub, log(55), 0.20), 37), 104),
  EGA = pmin(pmax(round(stats::rnorm(n_sub, 24, 5)), 13), 36)
)
cohort_pq <- tibble::tibble(
  id   = seq_len(n_sub),
  WT   = pmin(pmax(stats::rlnorm(n_sub, log(54), 0.20), 35), 115),
  PARA = pmin(pmax(stats::rlnorm(n_sub, log(680), 1.6), 5), 355400)
)
```

```{r cohort-sim}
set.seed(20240102)
ev_aq <- dplyr::bind_rows(lapply(
  seq_len(nrow(cohort_aq)),
  \(i) make_events_aq(cohort_aq$id[i], cohort_aq$WT[i], cohort_aq$EGA[i])
))
sim_aq <- rxode2::rxSolve(mod_aq, ev_aq, keep = c("WT", "EGA"),
                          useLinCmt = FALSE) |>
  as.data.frame()

set.seed(20240103)
ev_pq <- dplyr::bind_rows(lapply(
  seq_len(nrow(cohort_pq)),
  \(i) make_events_pq(cohort_pq$id[i], cohort_pq$WT[i], cohort_pq$PARA[i])
))
sim_pq <- rxode2::rxSolve(mod_pq, ev_pq, keep = c("WT", "PARA")) |>
  as.data.frame()
```

```{r pknca-build}
# One long concentration frame, with the analyte as the treatment grouping
# variable so PKNCA rolls its summaries up per analyte. The filter is
# !is.na(Cc) only, so the time-zero record is retained.
#
# Amodiaquine gets its own 240 h window. The shared grid runs to 150 days so
# that piperaquine's ~17-day terminal phase is characterised, but amodiaquine
# (t1/2 ~26 h) has decayed below any plausible quantification limit -- and
# into the solver's noise floor -- within a few days. Feeding PKNCA the full
# record makes its automatic lambda-z search fit that noise: half-life comes
# back at 67 h instead of 26 h and aucinf.obs fails outright for 145 of the
# 150 subjects. Truncating at 240 h (about 9 half-lives) is also the
# realistic analytical window: 79.2% of the trial's amodiaquine samples were
# below the LLOQ (Ding 2024 Results).
aq_nca_window_h <- 240

conc_all <- dplyr::bind_rows(
  sim_aq |>
    dplyr::filter(time <= aq_nca_window_h) |>
    dplyr::transmute(id, time, Cc = Cc,      treatment = "Amodiaquine"),
  sim_aq |> dplyr::transmute(id, time, Cc = Cc_deaq, treatment = "Desethylamodiaquine"),
  sim_pq |> dplyr::transmute(id, time, Cc = Cc,      treatment = "Piperaquine")
) |>
  dplyr::filter(!is.na(Cc))

stopifnot(sum(conc_all$time == 0) == 3 * n_sub)   # time-zero record present

dose_all <- dplyr::bind_rows(
  tidyr::expand_grid(id = cohort_aq$id, time = c(0, 24, 48),
                     amt = dose_aq_mg, treatment = c("Amodiaquine", "Desethylamodiaquine")),
  tidyr::expand_grid(id = cohort_pq$id, time = c(0, 24, 48),
                     amt = dose_pq_mg, treatment = "Piperaquine")
)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(conc_all), Cc ~ time | treatment + id,
                             concu = "ng/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_all), amt ~ time | treatment + id,
                             doseu = "mg")

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE, clast.obs = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r pknca-summary}
nca_tbl <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(
    Median = stats::median(PPORRES, na.rm = TRUE),
    `P5`   = stats::quantile(PPORRES, 0.05, na.rm = TRUE),
    `P95`  = stats::quantile(PPORRES, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

nca_tbl |>
  dplyr::rename("Analyte" = treatment, "NCA parameter" = PPTESTCD) |>
  knitr::kable(digits = 2, caption = paste(
    "PKNCA non-compartmental analysis of the simulated cohorts",
    "(n = 150 per arm). Concentrations in ng/mL, times in h."))
```

## Comparison against the published values

Ding 2024 reports AUC0-inf and terminal half-life for all three analytes,
and the day-7 concentration for desethylamodiaquine and piperaquine. Cmax
and Tmax are not reported by the paper and are therefore shown above but not
compared here.

```{r nca-comparison}
sim_for_cmp <- nca_tbl |>
  dplyr::filter(PPTESTCD %in% c("aucinf.obs", "half.life")) |>
  dplyr::transmute(treatment, PPTESTCD, PPORRES = Median)

ref_for_cmp <- tibble::tribble(
  ~treatment,             ~aucinf.obs,  ~half.life,
  "Amodiaquine",          283,          26.3,
  "Desethylamodiaquine",  45200,        14.1 * 24,
  "Piperaquine",          31600,        17.0 * 24
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = as.data.frame(sim_for_cmp),
  reference = as.data.frame(ref_for_cmp),
  by = "treatment",
  units = c(aucinf.obs = "h*ng/mL", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(cmp, digits = 2, caption = paste(
  "Simulated (cohort median) versus Ding 2024 published values.",
  attr(cmp, "footnote")))
```

```{r nca-assert}
pct <- suppressWarnings(as.numeric(gsub("[^0-9.eE+-]", "", cmp$`% diff`)))
pct <- pct[is.finite(pct)]
stopifnot(
  length(pct) == 6,
  # Every subject must yield a computable AUC and half-life. This is the
  # check that catches a mis-specified NCA window: before amodiaquine was
  # given its own 240 h window, aucinf.obs was NA for 145 of 150 subjects
  # and the group median was computed from the surviving 5.
  !any(is.na(nca_tbl$Median)),
  # Cohort-level check: assert on the centre and a robust envelope, never on
  # the extreme of a random cohort (see the repo note on cohort assertions).
  abs(stats::median(pct)) < 5,
  stats::quantile(abs(pct), 0.9) < 12
)
```

All six comparisons agree with the published values, and the simulated
5th-95th percentile ranges bracket the published 95% intervals (for
example, amodiaquine AUC0-inf 168-512 against a published 182-432).

The cohort medians sit slightly further from the published values than the
typical-value table above, which is expected: the published secondary
parameters are medians of post-hoc empirical Bayes estimates from a cohort
whose etas are heavily shrunk (70-74% shrinkage on the desethylamodiaquine
disposition etas, 73% on piperaquine `Vc/F`), whereas this cohort draws
unshrunk etas from the full `omega`. The very large `Vc/F` variability
(205% CV for desethylamodiaquine, 112% for piperaquine) widens the
half-life distribution substantially without moving AUC, which depends only
on clearance and bioavailability. The piperaquine AUC is the largest
discrepancy because it additionally integrates over the simulated baseline
parasitaemia distribution, which is an approximation of a covariate the
paper reports only as a median and range.

# Assumptions and deviations

1. **Allometric reference weight of 70 kg.** Equations 2 and 3 write the
   normalising constant generically as `BW_median`, but the footnotes of
   both Table 2 and Table 3 state that the population estimates are given
   for a typical pregnant woman weighing 70 kg, the Discussion compares the
   piperaquine `CL/F` of 69.9 L/h directly against a literature value "for a
   typical 70-kg adult patient", and the in-silico simulations use 70 kg
   hypothetical patients. 70 kg is therefore encoded. The typical-value
   table above shows a cohort-median reference (55 / 54 kg) would miss the
   published AUCs by 12-14%.

2. **Molar conversion on the amodiaquine-to-desethylamodiaquine flux.** The
   Methods state complete bioconversion but do not print a conversion
   factor. The model applies `MW_DEAQ / MW_AQ = 327.81 / 355.85 = 0.9212`,
   matching the sibling WWARN model `Ali_2018_amodiaquine`, which is
   reference 29 of this paper and states the correction explicitly. The
   paper's own numbers support it: the published `AUC_DEAQ / AUC_AQ` ratio
   of 45,200 / 283 = 159.7 back-solves a factor of 0.902 (this ratio
   cancels the dose and the bioavailability entirely), against 0.921 for the
   molar reading and 1.0 for a mass-for-mass reading.

3. **Amodiaquine absorption chain: `ktr = 2 / MTT` with `ka` as a separate
   final step.** Table 2 fixes both `Ka = 0.589 1/h` and `MTT = 0.236 h`
   with 2 transit compartments. These two numbers cannot both belong to a
   pure Savic chain, in which `ka` and `ktr` are the same rate (that would
   require `ka = 3 / 0.236 = 12.7`, not 0.589). Since `MTT` (0.236 h) is
   shorter than the mean time of the single `ka` step (1/0.589 = 1.70 h),
   `MTT` provably excludes the `ka` step -- a mean transit time cannot be
   smaller than one of the mean times it sums. `MTT` therefore spans only
   the two transit compartments, giving `ktr = 2 / MTT`, with `ka` governing
   the final transfer into the central compartment and a mean absorption
   time of 1.93 h. The sibling `Ali_2018_amodiaquine` model, which fixes the
   identical `Ka` / `MTT` pair, instead uses `ktr = 3 / MTT` and leaves `ka`
   unused; that reading discards a reported parameter. No published quantity
   discriminates between the two -- AUC, both half-lives and the day-7
   concentration are all insensitive to the absorption structure, and the
   paper reports no amodiaquine Cmax or Tmax -- so this choice is visible
   only in the unreported amodiaquine peak.

4. **Piperaquine absorption chain: `ktr = 3 / MTT`.** No separate `ka` is
   reported for piperaquine, so `ka = ktr` and the standard Savic
   convention applies over the three transitions
   `depot -> transit1 -> transit2 -> central`. This matches
   `Hoglund_2017_piperaquine`, which is reference 28 of this paper and the
   source of the fixed `MTT = 2.11 h`.

5. **Residual error reported as a variance.** The Table 2 and Table 3
   footnotes state "RUV is the residual error variance", so the tabulated
   `0.267` / `0.122` / `0.222` are variances and the model encodes
   `propSd = sqrt(RUV)`. The Methods describe an additive error on
   log-transformed concentrations, which is equivalent to a proportional
   error in linear concentration space. This is the same convention as the
   sibling `Hoglund_2017_piperaquine`.

6. **Base of the parasitaemia logarithm.** The Table 3 footnote writes
   `[1 + theta x (log(parasitemia) - 2.83)]` without stating the base. It is
   base 10: the cohort median is 680 parasites/uL and `log10(680) = 2.83`,
   and the Results simulations use "a baseline parasitemia of 676
   parasites/uL", which is `10^2.83 = 676.1`. `max(PARA, 1)` gating is
   applied because the Table 1 minimum is 0 parasites/uL.

7. **Text-versus-table discrepancies, resolved in favour of the
   final-estimates tables.** The Results text gives an 11.6% decrease in
   piperaquine bioavailability per log unit of parasitaemia while Table 3
   gives -11.9%; the abstract gives a 1.25%/week gestational-age effect
   while both the Results text and Table 2 give 1.28%/week. The Table 2 and
   Table 3 values are used.

8. **Gestational age is not extrapolable to the non-pregnant anchor.** The
   covariate register defines `EGA = 0` as the non-pregnant reference, but
   this model's linear form would give `F = 0.69` there. The cohort range is
   13-36 weeks and the relationship is only supported inside it; no
   non-pregnant women were enrolled.

9. **Virtual-cohort covariate distributions are approximations.** Table 1
   reports only medians and ranges, not distributional shapes or
   correlations. Weight and parasitaemia are drawn as truncated log-normals
   and gestational age as a truncated normal, each matched to the published
   median and clipped to the published range. Weight and gestational age are
   drawn independently, which the paper gives no basis to correlate.

10. **`deaq` metabolite suffix registered.** `central_deaq`,
    `peripheral1_deaq` and `propSd_deaq` follow the standard
    `<canonical>_<metab>` pattern and match the already-shipped
    `Ali_2018_amodiaquine`, which models the identical parent-metabolite
    pair. The suffix was not previously in the canonical register, so both
    files raised `checkModelConventions()` warnings; `deaq` is registered in
    `inst/references/compartment-names.md` alongside this extraction, which
    clears the warnings on both.

11. **Supplementary figures not used.** Figures S1-S6 (structure diagrams,
    goodness-of-fit plots, prediction-corrected VPCs) were not on disk. They
    contain no parameter values; Tables 1-3 in the main article are complete
    for both models.
