---
title: "Dose-response of biologics in psoriasis (Dodds 2013)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dose-response of biologics in psoriasis (Dodds 2013)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Dodds_2013_psoriasis_biologics_mbma"))$reference`
* Article: <https://doi.org/10.1038/psp.2013.32> (open access, CC BY-NC-ND 3.0)
* Supplement: the paper cites a Supplementary Table S1 and Supplementary
  Methods. Neither is required to reproduce the model: Table S1 is described in
  the text as giving "the true values of maximal response, ED50 and placebo
  response used in CTS", all of which are also printed in Table 1 and its
  footnote, and the Supplementary Methods describe the *estimation* step of the
  trial simulation rather than the dose-response model encoded here.

This is a model-based meta-analysis (MBMA) of dose-response, not a population PK
model. It has **no PK layer, no ODE states, and no time axis**: the source
predicts a single PASI observation per subject at the end of a 12-week window
following one dose. It does, however, carry a **subject-level** residual, which
distinguishes it from a per-arm MBMA such as
`modellib("Mandema_2011_biologicDMARDs_mbma")` by the same group.

Because there is no PK, there is nothing for `PKNCA` to integrate: no
concentration-time profile exists, so `Cmax`, `Tmax`, `AUC` and half-life are
all undefined for this model. The validation below therefore replaces the usual
NCA section with exact reproduction of the source's own printed dose-response
numbers, plus a check that the simulated subject-level spread matches the
source's variance equations.

# Population

Two different populations are involved and should not be conflated.

**The dose-response parameters** come from a model-based meta-analysis of
publicly available data from 27 randomized controlled trials of approved
compounds in psoriasis and psoriatic rheumatoid arthritis, covering anti-tumour
necrosis factor agents, ustekinumab and methotrexate. Trials were found by
PubMed search plus review of previous meta-analyses, clinicaltrials.gov,
conference abstracts and corporate websites. The meta-analysis jointly modelled
the probability of achieving PASI 50, 75 and 90 and the mean percent improvement
in PASI score; **only the percent-improvement arm is encoded here**. The source
publishes no subject count, no demographic table and no per-trial listing for
this database.

**The subject-level variance model** was calibrated separately, to reproduce the
between-subject variability of percent PASI improvement observed in the
PHOENIX 1 and PHOENIX 2 ustekinumab trials.

The simulated first-in-patient trials that the paper is actually about were 16
subjects each, 9,999 replicates per design per compound, run in NONMEM 7.2.
Those trial *designs* are not part of this model; the model is the truth
generator they were simulated from.

```{r population}
pop <- rxode2::rxode(readModelDb("Dodds_2013_psoriasis_biologics_mbma"))$population
tibble::tibble(Field = names(pop), Value = vapply(pop, as.character, character(1))) |>
  knitr::kable(caption = "Population metadata recorded with the model.")
```

# Model structure

The source writes the model in the log domain so that percent improvement in
PASI is bounded above by 100% on the real scale (Methods, "Clinical trial
simulation"):

$$
\mathrm{LPASI} = \log\!\left(\frac{\mathrm{PASI}_{\mathrm{treatment}}}{\mathrm{PASI}_{\mathrm{base}}}\right)
 = E_0 + \frac{E_{\max}\cdot \mathrm{Dose}}{ED_{50} + \mathrm{Dose}}
\qquad (\text{Eq. }1)
$$

with a subject-level residual whose standard deviation is **linear in the
prediction** rather than constant:

$$
Y = \mathrm{LPASI} + (s_{\mathrm{ERR}}\cdot\mathrm{LPASI} + i_{\mathrm{ERR}})\cdot\varepsilon,
\qquad \varepsilon \sim N(0,1),
\qquad s_{\mathrm{ERR}} = -0.32,\; i_{\mathrm{ERR}} = 0.30
\qquad (\text{Eqs. }2\text{--}3)
$$

`LPASI` is negative throughout (PASI falls under treatment), so the negative
`sERR` makes the residual standard deviation *grow* with response. The source
notes this explicitly: "the between-subject variability decreases with
increasing response (but not as much as expected with a simple additive error
model in the log domain), which motivated this choice of variance model."

`E0`, `Emax` and `ED50` are shared across compounds except for `ED50`: the three
marketed antibodies were found to have a **common maximum effect** and to differ
only in potency.

## Encoding note: how Eq. 2 maps onto rxode2

rxode2's `combined1()` variance model computes the residual standard deviation
as `addSd + propSd * prediction` on the *signed* prediction, which is exactly
the shape of Eq. 2 -- but rxode2 requires `propSd > 0`, and Eq. 2's slope is
negative. The model block therefore predicts the log PASI **reduction**
`pasiLogDrop = -LPASI`, which is positive over the whole dose range, so the
slope becomes `+0.32`:

$$
SD = 0.30 + 0.32\cdot(-\mathrm{LPASI}) = 0.30 - 0.32\cdot \mathrm{LPASI}
   = i_{\mathrm{ERR}} + s_{\mathrm{ERR}}\cdot \mathrm{LPASI}
$$

Because `eps` is symmetric about zero, negating the endpoint leaves the
distribution of the response unchanged: the encoded model is Eq. 2 exactly, not
an approximation. The check under "Subject-level variability" below verifies the
simulated standard deviation against `iERR + sERR*LPASI` dose by dose.

# Recovering the log-domain parameters

The source reports its dose-response on the **real** scale (Table 1 gives the
maximal absolute difference from placebo in PASI percentage points; the table
footnote gives the placebo response) while Eq. 1 is written in the **log**
domain. The two log-domain values in `ini()` are therefore obtained by inverting
the source's own definitions.

```{r recover}
placebo_pct <- 0.095    # Table 1 footnote: "Placebo response (E0) is 9.5% for all compounds."
maxdiff_pct <- 0.823    # Table 1: maximal absolute difference from placebo, marketed compounds

# Percent improvement = 1 - exp(LPASI), so a 9.5% placebo response is a PASI
# ratio of 0.905.
e0_rec <- log(1 - placebo_pct)

# The maximal ABSOLUTE difference from placebo is taken on the real scale:
#   exp(e0) - exp(e0 + emax) = 0.823
emax_rec <- log(1 - maxdiff_pct / (1 - placebo_pct))

ini_df <- rxode2::rxode(readModelDb("Dodds_2013_psoriasis_biologics_mbma"))$iniDf

tibble::tibble(
  Parameter   = c("e0", "emax"),
  Recovered   = c(e0_rec, emax_rec),
  `In ini()`  = c(ini_df$est[ini_df$name == "e0"],
                  ini_df$est[ini_df$name == "emax"])
) |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 7))) |>
  knitr::kable(caption = "Log-domain parameters recovered from the real-scale table.")

# Deterministic arithmetic, so an exact check is appropriate here.
stopifnot(
  abs(e0_rec   - ini_df$est[ini_df$name == "e0"])   < 1e-7,
  abs(emax_rec - ini_df$est[ini_df$name == "emax"]) < 1e-7
)
```

The inversion is confirmed below against three printed numbers that were *not*
used to derive it: the absolute difference from placebo at 700 mg for each of
the three marketed antibodies.

# Source trace

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Item, ~`Source location in Dodds 2013`,
  "Eq. 1, LPASI = E0 + Emax*Dose/(ED50 + Dose)", "Methods, 'Clinical trial simulation', Eq. 1",
  "Eqs. 2-3, subject-level variance model", "Methods, 'Clinical trial simulation', Eqs. 2 and 3",
  "sERR = -0.32, iERR = 0.30", "Methods, Eq. 3 (printed inline)",
  "e0 = log(1 - 0.095)", "Table 1 footnote: 'Placebo response (E0) is 9.5% for all compounds.'",
  "emax = log(1 - 0.823/0.905)", "Table 1, 'Maximal absolute difference from placebo PASI % change (%)' = 82.3 for all marketed compounds; Results, 'Test compound selection'",
  "ED50 adalimumab = 16.9 mg", "Table 1, adalimumab row; Results: 'ED50 = 16.9, 45.5, and 13.9 mg for adalimumab, golimumab, and ustekinumab, respectively'",
  "ED50 golimumab = 45.5 mg", "Table 1, golimumab row; same Results sentence",
  "ED50 ustekinumab = 13.9 mg", "Table 1, ustekinumab row; same Results sentence",
  "Twofold ED50 acceptance windows", "Table 1, 'Estimated ED50 within' column",
  "Absolute difference from placebo at 700 mg (81.8 / 81.0 / 81.9%)", "Results, 'Design performance metrics -> G/NG criteria', second paragraph",
  "Go/No-Go criterion: > 50 percentage points at 700 mg", "Results, 'G/NG criteria'; Abstract",
  "Simulated dose levels 0, 21, 70, 210, 700 mg", "Results, 'Distributed trial designs'; Table 2; Figure 1 caption",
  "70 mg/mL presentation, 0.3 / 1 / 3 / 10 mL", "Results, 'Distributed trial designs'",
  "PHOENIX 2 comparison at 45 and 90 mg ustekinumab", "Figure 3 and its caption; Methods, 'Clinical trial simulation'",
  "27 pooled trials; covariates screened", "Methods, 'MBMA of biologic treatments for psoriasis'",
  "Hypothetical compounds (excluded here)", "Table 1, 'No-Go examples' and 'Go examples' rows"
) |>
  knitr::kable(caption = "Provenance of every equation and every ini() value.")
```

# Virtual cohort

There is no PK profile to sample and no time axis. The "cohort" is a grid of
(compound, dose) combinations: each row is one subject assigned one single dose
of one antibody, exactly as a subject in the source's simulated first-in-patient
trial would be.

```{r cohort}
mod <- rxode2::rxode(readModelDb("Dodds_2013_psoriasis_biologics_mbma"))

dose_cols <- c(adalimumab  = "CONMED_ADALIMUMAB_DOSE",
               golimumab   = "CONMED_GOLIMUMAB_DOSE",
               ustekinumab = "CONMED_USTEKINUMAB_DOSE")

# Table 1 ED50 values, used only to build the dose grid and to check the
# twofold acceptance windows -- not to parameterise anything.
ed50_pub <- c(adalimumab = 16.9, golimumab = 45.5, ustekinumab = 13.9)

# One row per (compound, dose). Every dose column is 0 except the one named,
# which is how a placebo subject is encoded (all three columns 0).
make_rows <- function(label, drug, doses) {
  out <- tibble::tibble(
    arm  = label,
    drug = drug,
    dose = doses,
    time = 0,
    evid = 0,
    CONMED_ADALIMUMAB_DOSE  = 0,
    CONMED_GOLIMUMAB_DOSE   = 0,
    CONMED_USTEKINUMAB_DOSE = 0
  )
  if (!is.na(drug)) out[[dose_cols[[drug]]]] <- doses
  out
}

# 1. Placebo: every dose column 0.
rows_placebo <- make_rows("placebo", NA_character_, 0)

# 2. The five dose levels the source simulated, for each marketed antibody.
sim_doses <- c(0, 21, 70, 210, 700)
rows_levels <- dplyr::bind_rows(lapply(names(dose_cols), function(d) {
  make_rows(paste0("level:", d), d, sim_doses)
}))

# 3. A very large dose per antibody, to read the asymptote (Emax) off the model.
rows_asym <- dplyr::bind_rows(lapply(names(dose_cols), function(d) {
  make_rows(paste0("asym:", d), d, 1e9)
}))

# 4. Each antibody AT its published ED50, and at the twofold window edges.
rows_ed50 <- dplyr::bind_rows(lapply(names(dose_cols), function(d) {
  make_rows(paste0("ed50:", d), d, ed50_pub[[d]])
}))

# 5. A smooth dose grid for Figure 1.
n_grid <- 120L
grid_doses <- c(0, exp(seq(log(0.5), log(1500), length.out = n_grid - 1L)))
rows_grid <- dplyr::bind_rows(lapply(names(dose_cols), function(d) {
  make_rows(paste0("grid:", d), d, grid_doses)
}))

# 6. The two licensed ustekinumab doses compared against PHOENIX 2 in Figure 3.
rows_phoenix <- make_rows("phoenix:ustekinumab", "ustekinumab", c(45, 90))

arms <- dplyr::bind_rows(rows_placebo, rows_levels, rows_asym, rows_ed50,
                         rows_grid, rows_phoenix) |>
  dplyr::mutate(id = dplyr::row_number(), .before = 1)

cat("rows simulated:", nrow(arms), "\n")
```

# Simulation

The structural model is deterministic -- there are no etas -- so a single solve
returns the typical-value prediction for every row. (`omega = NA` is *not*
passed: the model declares no random effects, and rxode2 errors on `omega = NA`
when there is no IIV to suppress.)

```{r simulate}
typsim <- rxode2::rxSolve(
  mod,
  events = as.data.frame(dplyr::select(arms, -"drug", -"dose")),
  keep = "arm",
  returnType = "data.frame"
) |>
  tibble::as_tibble() |>
  dplyr::mutate(id = as.integer(as.character(id))) |>
  dplyr::left_join(dplyr::select(arms, id, drug, dose), by = "id")

stopifnot(nrow(typsim) == nrow(arms))

get_arm <- function(label) typsim[typsim$arm == label, ]
```

## The placebo response is reproduced exactly

Setting all three dose columns to 0 collapses the drug effect and leaves `e0`.

```{r placebo-check}
plc <- get_arm("placebo")
cat(sprintf("placebo percent PASI improvement: %.4f%% (Table 1 footnote: 9.5%%)\n",
            plc$pasiPctImprove))

# Deterministic: this is arithmetic on a fixed() parameter, so assert tightly.
stopifnot(abs(plc$pasiPctImprove - 9.5) < 1e-6,
          abs(plc$pasiPctImproveDrug - 0) < 1e-9)
```

## The maximum effect is 82.3 percentage points for all three antibodies

At a dose far above every `ED50`, the dose fraction saturates and the absolute
difference from placebo must equal the single shared `Emax` expressed on the
real scale.

```{r asymptote-check}
asym <- typsim |>
  dplyr::filter(grepl("^asym:", arm)) |>
  dplyr::transmute(Compound = drug,
                   `Maximal difference from placebo (pp)` = pasiPctImproveDrug,
                   `Published (Table 1, pp)` = 82.3)

asym |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 4))) |>
  knitr::kable(caption = "Asymptotic drug effect versus Table 1.")

stopifnot(max(abs(asym$`Maximal difference from placebo (pp)` - 82.3)) < 1e-4)
```

## Effect at 700 mg reproduces three numbers not used to build the model

This is the strongest available check. `e0` and `emax` were recovered from the
9.5% placebo response and the 82.3-point asymptote; the `ED50` values came from
Table 1. None of that used the Results narrative, which separately states that
adalimumab, golimumab and ustekinumab "demonstrate an 81.8, 81.0, and 81.9%
absolute difference between drug effect at 700 mg and placebo effect,
respectively". If any of the five parameters were mis-transcribed, or the
log/real-domain inversion were wrong, these three numbers would not land.

```{r seven-hundred-check}
published_700 <- c(adalimumab = 81.8, golimumab = 81.0, ustekinumab = 81.9)

at700 <- typsim |>
  dplyr::filter(grepl("^level:", arm), dose == 700) |>
  dplyr::transmute(
    Compound              = drug,
    `ED50 (mg)`           = ed50_pub[drug],
    `Simulated (pp)`      = pasiPctImproveDrug,
    `Published (pp)`      = published_700[drug],
    `Difference (pp)`     = pasiPctImproveDrug - published_700[drug],
    `Go/No-Go (> 50 pp)`  = ifelse(pasiPctImproveDrug > 50, "G", "NG"),
    `Published G/NG`      = "G"
  )

at700 |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Absolute PASI improvement over placebo at the maximum feasible 700 mg dose.")

# Deterministic prediction versus a number printed to one decimal place, so the
# tolerance is the printing precision (0.05 pp), not a simulation tolerance.
stopifnot(
  max(abs(at700$`Difference (pp)`)) < 0.05,
  all(at700$`Go/No-Go (> 50 pp)` == at700$`Published G/NG`)
)
```

All three marketed compounds clear the source's Go/No-Go threshold, matching the
"Correct G/NG" column of Table 1.

## The twofold ED50 acceptance windows reproduce Table 1

Table 1's "Estimated ED50 within" column is twofold either side of the true
`ED50`; reproducing it is an independent check that the `ED50` values were read
off the correct rows.

```{r ed50-window-check}
published_window <- tibble::tribble(
  ~drug,         ~lo,   ~hi,
  "adalimumab",  8.50,  33.8,
  "golimumab",  22.8,   91.0,
  "ustekinumab", 6.90,  27.7
)

window_check <- published_window |>
  dplyr::mutate(
    `ED50 (mg)`        = ed50_pub[drug],
    `ED50/2 (mg)`      = `ED50 (mg)` / 2,
    `ED50*2 (mg)`      = `ED50 (mg)` * 2
  ) |>
  dplyr::rename(Compound = drug, `Published lower (mg)` = lo,
                `Published upper (mg)` = hi)

window_check |>
  knitr::kable(caption = "Twofold ED50 windows: computed versus Table 1 as printed.")

# Table 1 prints the windows to three significant figures and rounds outward,
# so agreement is to the printing precision rather than exact.
stopifnot(
  max(abs(window_check$`ED50/2 (mg)` - window_check$`Published lower (mg)`)) < 0.06,
  max(abs(window_check$`ED50*2 (mg)` - window_check$`Published upper (mg)`)) < 0.11
)
```

Also confirmed: at `Dose = ED50` the drug term is exactly half of `Emax` in the
log domain, which is the definition the source gives for `ED50`.

```{r ed50-half-check}
emax_log <- ini_df$est[ini_df$name == "emax"]
half <- typsim |>
  dplyr::filter(grepl("^ed50:", arm)) |>
  dplyr::transmute(Compound = drug,
                   `Drug effect at ED50 / Emax` = (lpasi - plc$lpasi) / emax_log)

half |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 10))) |>
  knitr::kable(caption = "At Dose = ED50 the drug effect is exactly half of Emax.")

stopifnot(max(abs(half$`Drug effect at ED50 / Emax` - 0.5)) < 1e-9)
```

# Replicate Figure 1: dose-response of the marketed compounds

Replicates the solid lines of Figure 1 of Dodds 2013 (the marketed biologics
adalimumab, golimumab and ustekinumab), with the source's simulated dose range
shaded. The hypothetical compounds shown in the source's dashed lines are not
part of this model -- see "Assumptions and deviations".

```{r figure1, fig.cap = "Replicates the solid lines of Figure 1 of Dodds 2013: expected typical percent reduction in PASI versus dose for the three marketed biologics. The shaded band is the 0-700 mg range used in the trial simulation."}
grid_sim <- typsim |> dplyr::filter(grepl("^grid:", arm))

ggplot(grid_sim, aes(x = dose, y = pasiPctImprove, colour = drug)) +
  annotate("rect", xmin = 0, xmax = 700, ymin = -Inf, ymax = Inf,
           fill = "steelblue", alpha = 0.12) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 100 * (1 - exp(plc$lpasi)), linetype = "dotted") +
  scale_x_log10() +
  labs(x = "Single dose (mg, log scale)",
       y = "Typical percent improvement in PASI (%)",
       colour = "Compound") +
  theme_bw()
```

# Subject-level variability (Eqs. 2-3)

The source adds one PASI observation per subject with a standard deviation that
is linear in the prediction. Simulating a cohort at each dose level and
comparing the empirical standard deviation against `iERR + sERR*LPASI` checks
that the negated-endpoint encoding reproduces Eq. 2.

```{r residual-check}
n_per_arm <- 200L   # per the skill's cap; SE of an SD at n = 200 is about 5%

resid_arms <- dplyr::bind_rows(lapply(sim_doses, function(d) {
  r <- make_rows("resid", "ustekinumab", rep(d, n_per_arm))
  r$dose_level <- d
  r
})) |>
  dplyr::mutate(id = dplyr::row_number(), .before = 1)

rxode2::rxSetSeed(20130724)
resid_sim <- rxode2::rxSolve(
  mod,
  events = as.data.frame(dplyr::select(resid_arms, -"arm", -"drug", -"dose",
                                       -"dose_level")),
  returnType = "data.frame"
) |>
  tibble::as_tibble() |>
  dplyr::mutate(id = as.integer(as.character(id))) |>
  dplyr::left_join(dplyr::select(resid_arms, id, dose_level), by = "id")

resid_check <- resid_sim |>
  dplyr::group_by(dose_level) |>
  dplyr::summarise(
    LPASI          = unique(round(lpasi, 10)),
    `Simulated SD` = stats::sd(sim),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    `Eq. 2-3 SD`   = 0.30 + (-0.32) * LPASI,
    `Ratio`        = `Simulated SD` / `Eq. 2-3 SD`
  ) |>
  dplyr::rename(`Dose (mg)` = dose_level)

resid_check |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 4))) |>
  knitr::kable(caption = "Empirical subject-level SD of the simulated log PASI response versus Eqs. 2-3.")

# Cohort-derived: the bound must admit sampling noise, not one machine's draw.
# With n = 200 per dose the relative standard error of an SD is about 5%, and
# rxode2's RNG stream is partitioned per solver thread, so a CI runner draws a
# different cohort. A 25% band is about five standard errors -- wide enough to
# hold for any thread count, and still narrow enough to go red if the slope or
# intercept of the variance model were mis-transcribed (dropping sERR entirely
# would put the ratio at 0.28 at 700 mg; dropping iERR would put it at 1.4 at
# placebo).
# Realised max|Ratio - 1| of 0.077 / 0.080 / 0.075 at 2 / 8 / 16 solver threads,
# so 0.25 sits about three times outside the observed spread. Do not tighten it
# back to any single run's value.
stopifnot(max(abs(resid_check$Ratio - 1)) < 0.25)
```

# Replicate Figure 3: simulated response versus PHOENIX 2

Figure 3 of Dodds 2013 overlays the simulated arithmetic mean and standard
deviation of percent PASI improvement for placebo, 45 mg ustekinumab and 90 mg
ustekinumab on the observed PHOENIX 2 data. The source prints no numeric values
for either series, so this section reproduces the *simulated* (red) series only;
the observed (blue) series would have to be digitised from the figure and is not
reproduced here.

Note that percent improvement is a non-linear transform of the simulated log
response, so the cohort **mean** percent improvement sits below the
**typical-value** percent improvement. Both are shown.

```{r figure3, fig.cap = "Replicates the simulated (red) series of Figure 3 of Dodds 2013: arithmetic mean and standard deviation of percent PASI improvement for placebo and for 45 and 90 mg ustekinumab. The observed PHOENIX 2 series is not reproduced (the source prints no numeric values for it)."}
fig3_doses <- c(0, 45, 90)
fig3_arms <- dplyr::bind_rows(lapply(fig3_doses, function(d) {
  r <- make_rows("fig3", "ustekinumab", rep(d, n_per_arm))
  r$dose_level <- d
  r
})) |>
  dplyr::mutate(id = dplyr::row_number(), .before = 1)

rxode2::rxSetSeed(20130724)
fig3_sim <- rxode2::rxSolve(
  mod,
  events = as.data.frame(dplyr::select(fig3_arms, -"arm", -"drug", -"dose",
                                       -"dose_level")),
  returnType = "data.frame"
) |>
  tibble::as_tibble() |>
  dplyr::mutate(id = as.integer(as.character(id))) |>
  dplyr::left_join(dplyr::select(fig3_arms, id, dose_level), by = "id") |>
  # sim is the log PASI REDUCTION, so percent improvement is 1 - exp(-sim).
  dplyr::mutate(pct_improve = 100 * (1 - exp(-sim)))

fig3_summary <- fig3_sim |>
  dplyr::group_by(dose_level) |>
  dplyr::summarise(
    `Typical (%)` = 100 * (1 - exp(unique(round(lpasi, 10)))),
    `Mean (%)`    = mean(pct_improve),
    `SD (%)`      = stats::sd(pct_improve),
    .groups = "drop"
  ) |>
  dplyr::rename(`Ustekinumab dose (mg)` = dose_level)

fig3_summary |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable(caption = "Simulated percent PASI improvement, mean and SD (the red series of Figure 3).")

ggplot(fig3_summary, aes(x = factor(`Ustekinumab dose (mg)`), y = `Mean (%)`)) +
  geom_point(colour = "firebrick", size = 3) +
  geom_errorbar(aes(ymin = `Mean (%)` - `SD (%)`,
                    ymax = `Mean (%)` + `SD (%)`),
                width = 0.15, colour = "firebrick") +
  labs(x = "Ustekinumab single dose (mg)",
       y = "Percent improvement in PASI (%)") +
  theme_bw()
```

The qualitative feature the source calls out is reproduced: the standard
deviation of percent improvement **falls** as response rises (placebo is the
widest band), even though the standard deviation on the log scale grows. That
is the behaviour the linear variance model was chosen to produce.

```{r figure3-check}
# Trend, not step-by-step monotonicity or an exact value: this is a cohort
# statistic (pattern 12 of the known-failure catalogue). The realised placebo-
# minus-90 mg SD gap was 11.9 / 7.7 / 12.8 percentage points at 2 / 8 / 16
# solver threads, so a 3-point floor holds for any thread count while still
# going red if the variance model lost its dependence on the prediction.
stopifnot(
  fig3_summary$`SD (%)`[fig3_summary$`Ustekinumab dose (mg)` == 90] <
    fig3_summary$`SD (%)`[fig3_summary$`Ustekinumab dose (mg)` == 0] - 3,
  # The 45 and 90 mg typical values are deterministic and must bracket the
  # licensed-dose response the source simulated.
  all(fig3_summary$`Typical (%)`[fig3_summary$`Ustekinumab dose (mg)` > 0] > 80)
)
```

# Assumptions and deviations

* **Only the three marketed antibodies are extracted.** Table 1 of the source
  lists eight compounds: adalimumab, golimumab and ustekinumab (real, with
  meta-analysis-derived parameters) plus five hypothetical compounds
  (`discontinumab`, `mehmimab`, `cuspmimab`, `lowpomab`, `nopomab`) whose
  parameters were *hand-chosen* to stress-test the trial designs, not estimated
  from data. Shipping fictional drug names in a model library would be
  misleading, so they are excluded; this was an explicit operator decision
  (task `frompeople-1103`, sidecar request-001, answer B). Their parameters are
  fully recoverable from the same two identities used above if anyone needs
  them: `emax = log(1 - d/0.905)` for a maximal difference `d`, with the ED50
  read from Table 1.
* **The trial designs themselves are not modelled.** The source's contribution
  is a comparison of concentrated (placebo vs 700 mg only) and distributed
  (placebo plus 21 / 70 / 210 / 700 mg) first-in-patient designs across 9,999
  simulated 16-subject trials each. That is a simulation *study*, not a model;
  what is reusable, and what is extracted here, is the dose-response truth
  generator plus its subject-level variance model.
* **`e0` and `emax` are derived, not printed.** The source reports the
  dose-response on the real scale but writes Eq. 1 in the log domain, and never
  prints the log-domain values. Both were recovered by inverting the source's
  own definitions, as shown in "Recovering the log-domain parameters", and
  validated against three independent printed numbers.
* **The residual is encoded on the negated endpoint.** `pasiLogDrop = -LPASI`
  is used so rxode2's `combined1()` variance model can express Eq. 2's negative
  slope with a positive `propSd`. This is an exact re-parameterisation, not an
  approximation -- see the encoding note above and the verified check under
  "Subject-level variability".
* **The observation variable `pasiLogDrop` is not a canonical nlmixr2lib output
  name**, and `checkModelConventions()` raises one warning for it. There is no
  drug concentration to call `Cc` and no registered PD-output canonical for a
  PASI score. This follows the existing practice for score-type PD endpoints in
  the library (`walkDist` in `Hajjar_2018_DMD_6MWT`, `fev1` in
  `Wu_2014_FEV1_asthma`, `BMD` in `Plan_2012_bmd_fracture`).
* **`CONMED_<drug>_DOSE` columns carry a different dose metric here than in
  `Mandema_2011_biologicDMARDs_mbma`.** In that model they are per-administration
  doses on each drug's maintenance regimen (adalimumab 40 mg q2w); here they are
  **single-dose** milligram amounts. The two models' dose values must never be
  carried across. This is recorded in `covariateData` and in the covariate
  register entry.
* **No PKNCA validation.** There is no PK layer, no time axis and no
  concentration, so every NCA parameter is undefined for this model. The
  dose-response identity checks above take its place.
* **No between-study random effect is encoded.** The source reports none for the
  percent-improvement dose-response used in the simulation, and the variability
  it does specify (Eqs. 2-3) is explicitly subject-level, calibrated to
  PHOENIX 1 and 2. Nothing was invented to fill the gap.
* **Population demographics are unavailable.** The source publishes no subject
  count, no demographic table and no per-trial listing for its 27-trial
  meta-analysis database, so `n_subjects` and the demographic fields of
  `population` are absent rather than guessed.
* **Supplement not on disk.** Supplementary Table S1 and the Supplementary
  Methods were not retrieved. Per the source's own description, Table S1 repeats
  the maximal response, ED50 and placebo response already printed in Table 1,
  and the Supplementary Methods cover the estimation step of the trial
  simulation rather than the dose-response model. No parameter encoded here
  depends on them.
* **No errata found.** The article landing page and PubMed record show no
  correction, corrigendum or author notice for doi:10.1038/psp.2013.32.
