---
title: "Piperacillin (Dong 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Piperacillin (Dong 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
ui <- rxode2::rxode(readModelDb("Dong_2026_piperacillin"))
```

* Citation: `r ui$reference`
* Description: `r ui$description`
* Article: <https://doi.org/10.2147/DDDT.S551307>

Dong and colleagues fit a one-compartment model with first-order elimination to
117 steady-state plasma piperacillin concentrations from 42 critically ill
adults in a single Chinese intensive care unit. Clearance carries a
cystatin-C-based CKD-EPI estimated glomerular filtration rate and total body
weight; central volume carries serum albumin. All three covariates enter as
power functions centred on their cohort medians.

# Population

The cohort (Tables 1 and 2 of the source) was 42 ICU patients, 30 male and 12
female, with a median age of 59 years (IQR 50.25-75.5), median total body
weight 70 kg (IQR 65-75) and median BMI 24.22 kg/m^2 (IQR 22.41-26.12). Median
APACHE II score was 21.5 (IQR 19-26); 85.71% were mechanically ventilated and
85.71% were receiving a vasopressor. Pulmonary infection predominated (92.86%),
and *Pseudomonas aeruginosa* was the most common isolate (26.19%). Almost all
patients (92.86%) received 4 g piperacillin q8h by intravenous infusion.

Renal function is the striking feature of this cohort, and it is why the choice
of estimating equation matters. The median 2012 CKD-EPI **cystatin-C** eGFR --
the equation the final model uses -- is 46.56 mL/min/1.73 m^2 (IQR 26.88-68.26),
whereas the same patients' median **creatinine**-based estimates are roughly
twice as high (2021 CKD-EPI creatinine 93.32; abbreviated MDRD 107.30; MDRD CHN
112.63). The authors screened Cockcroft-Gault, MDRD, CKD-EPIcr, CKD-EPIcys and
CKD-EPIcr-cys and retained CKD-EPIcys because it gave the largest objective
function drop, reasoning that creatinine is confounded by muscle mass and
systemic inflammation in critical illness. **A user supplying real data must use
the cystatin-C equation**; substituting a creatinine-based eGFR for the same
patient would roughly double the covariate and inflate predicted clearance.

Patients on continuous renal replacement therapy were excluded, so the model
carries no information about extracorporeal clearance.

```{r population}
str(ui$population, max.level = 1)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Dong_2026_piperacillin.R`. The table below
collects them in one place.

A note on retrieval: the two final-model equations are set as **vector
graphics** on page 6 of the PDF and carry no text layer, so they are invisible
to `pdftotext` and absent from every text conversion of this paper. They were
read from a rendered image of the page. They are the only source for the three
centring constants, and -- as the omega-scale check below shows -- the only
thing that settles the scale of the inter-individual variability column.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL) | 6.48 L/h | Table 3, "CL (L/h)"; leading coefficient of the p. 6 CL equation |
| `lvc` (V) | 19 L | Table 3, "V (L)"; leading coefficient of the p. 6 V equation |
| `e_crcl_cl` | 0.615 | Table 3, "eGFR on CL"; exponent in the p. 6 CL equation |
| `e_wt_cl` | 1.13 | Table 3, "Total body weight on CL"; exponent in the p. 6 CL equation |
| `e_alb_vc` | 1.21 | Table 3, "Albumin on V"; exponent in the p. 6 V equation |
| eGFR centring constant | 46.56 mL/min/1.73 m^2 | p. 6 CL equation denominator; equals the Table 1 median for "2012 CKD-EPI CYS-C" |
| Weight centring constant | 70 kg | p. 6 CL equation denominator; equals the Table 1 median total body weight |
| Albumin centring constant | 34.8 g/L | p. 6 V equation denominator; equals the Table 1 median albumin |
| `etalcl` | 0.114 | p. 6 CL equation factor `e^0.114`; = 0.338^2 from Table 3 "IIV-CL (%)" 33.8 |
| `etalvc` | 0.069 | p. 6 V equation factor `e^0.069`; = 0.263^2 from Table 3 "IIV-V (%)" 26.3 |
| `propSd` | 0.177 | Table 3, "RSV_CV (%)" 17.7; footnote defines RSV_CV as proportional |
| `addSd` | 1.2 mg/L | Table 3, "RSV_SD (mg/L)" 1.2; footnote defines RSV_SD as additive |
| `d/dt(central)` (1-cmt, first-order) | n/a | Results, "Population Pharmacokinetics Model Development" |
| Combined residual error | n/a | Results: "residual variability were best described by a combined model" |

## Check 1 -- the scale of the inter-individual variability column

Table 3 labels the IIV rows `IIV-CL (%) = 33.8` and `IIV-V (%) = 26.3`. A bare
percentage is ambiguous between a log-scale standard deviation and a lognormal
coefficient of variation, and the two readings differ enough to matter. The
paper's own printed equations resolve it: they carry the literal factors
`e^0.114` on CL and `e^0.069` on V, which are the eta terms with the fitted
variance substituted for the random draw. Only the SD reading reproduces them.

```{r omega-scale}
iiv_pct <- c(CL = 33.8, V = 26.3)      # Table 3
printed  <- c(CL = 0.114, V = 0.069)   # p. 6 equation factors e^0.114 and e^0.069

omega_scale <- tibble::tibble(
  Parameter = names(iiv_pct),
  `Printed equation exponent` = printed,
  `SD reading: (IIV/100)^2` = round((iiv_pct / 100)^2, 4),
  `CV reading: log(1+(IIV/100)^2)` = round(log(1 + (iiv_pct / 100)^2), 4),
  `Encoded` = c(ui$omega["etalcl", "etalcl"], ui$omega["etalvc", "etalvc"])
)
knitr::kable(omega_scale, digits = 4,
             caption = "The printed equation exponents match the SD reading, not the CV reading.")

# Deterministic: the SD reading must reproduce the printed exponents to the
# 3 decimals they are printed at, and the CV reading must NOT.
stopifnot(
  all(abs(round((iiv_pct / 100)^2, 3) - printed) < 1e-9),
  all(abs(round(log(1 + (iiv_pct / 100)^2), 3) - printed) > 0),
  # and the model file encodes the printed values themselves
  abs(ui$omega["etalcl", "etalcl"] - 0.114) < 1e-12,
  abs(ui$omega["etalvc", "etalvc"] - 0.069) < 1e-12
)
```

Read the other way round, an omega^2 of 0.114 implies a lognormal CV of
`r round(100 * sqrt(exp(0.114) - 1), 1)`%, not the tabulated 33.8%. The
tabulated percentages are therefore `100 * omega`.

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the Table 1 marginal distributions: each covariate is drawn lognormally with the
published median as its median and the published interquartile range setting its
spread. The paper does not publish the joint covariate distribution it simulated
from, nor any correlation between eGFR, weight and albumin, so the three are
drawn independently -- this is an assumption, recorded again under *Assumptions
and deviations*.

```{r cohort}
# set.seed() seeds R's RNG, not rxode2's; rxode2 partitions its streams per
# solver thread, so a 2-core CI runner draws a different cohort than a
# 16-thread workstation. Every assertion below is written to hold for any
# cohort the model can produce (see known-vignette-failure-patterns.md #12).
set.seed(20260912)
rxode2::rxSetSeed(20260912)

n_sub <- 200L   # cap is 200 per arm
qn <- qnorm(0.75)

# Lognormal spread implied by a published median and IQR.
sd_from_iqr <- function(q25, q75) log(q75 / q25) / (2 * qn)

cohort <- tibble::tibble(
  id   = seq_len(n_sub),
  # Table 1: 2012 CKD-EPI CYS-C 46.56 (26.88-68.26) mL/min/1.73 m^2
  CRCL = 46.56 * exp(rnorm(n_sub, 0, sd_from_iqr(26.88, 68.26))),
  # Table 1: total body weight 70 (65-75) kg
  WT   = 70    * exp(rnorm(n_sub, 0, sd_from_iqr(65, 75))),
  # Table 1: albumin 34.8 (32.7-39.2) g/L
  ALB  = 34.8  * exp(rnorm(n_sub, 0, sd_from_iqr(32.7, 39.2)))
)

knitr::kable(
  cohort |>
    tidyr::pivot_longer(-id, names_to = "Covariate", values_to = "v") |>
    group_by(Covariate) |>
    summarise(Median = median(v), Q25 = quantile(v, .25), Q75 = quantile(v, .75),
              .groups = "drop") |>
    dplyr::rename("25th pctile" = Q25, "75th pctile" = Q75),
  digits = 2, caption = "Simulated cohort vs. the Table 1 medians (46.56, 70, 34.8) and IQRs."
)
```

```{r events-helper}
# Build an event table for one regimen across the whole cohort.
#   daily_mg : total daily dose in mg
#   tau      : dosing interval in h, or NA for continuous intravenous infusion
#   dur      : infusion duration in h for intermittent regimens
# Observations are written on the ODE state `central`; rxode2 returns the
# algebraic observable as a column at those rows. Naming an observable as a
# compartment instead would inject a slot and renumber the ODE states.
# `days` sets the dosing horizon before the sampled interval. It is 10 rather
# than the 2-3 that the typical subject needs, because the lognormal eGFR tail
# produces subjects with a clearance near 1.7 L/h and a half-life near 12 h;
# those subjects are still approaching steady state at 72 h, which shows up as
# a mass-balance residual in Check 3 rather than as any visible artefact.
make_events <- function(cohort, daily_mg, tau, dur = 0.5, days = 10,
                        obs_by = 0.02, label = NULL) {
  horizon <- 24 * days
  if (is.na(tau)) {
    ev <- rxode2::et(amt = daily_mg * days, dur = horizon, cmt = "central")
    win <- 8                                   # any window; Css is flat
  } else {
    ndose <- ceiling(horizon / tau)
    ev <- rxode2::et(amt = daily_mg * tau / 24, dur = dur,
                     ii = tau, addl = ndose - 1L, cmt = "central")
    win <- tau                                 # one full steady-state interval
  }
  # Sample the LAST complete interval inside the dosing horizon.
  t0 <- horizon - 24
  ev <- rxode2::et(ev, seq(t0, t0 + win, by = obs_by), cmt = "central")
  ev <- as.data.frame(ev)
  out <- merge(cbind(cohort, .k = 1L), cbind(ev, .k = 1L), by = ".k")
  out$.k <- NULL
  if (!is.null(label)) out$regimen <- label
  out[order(out$id, out$time), ]
}
```

# Simulation

```{r simulate}
mod <- readModelDb("Dong_2026_piperacillin")

# The clinical regimen: 4 g q8h as a 30-min infusion (92.86% of the cohort,
# Table 2; the paper's simulations standardise intermittent infusions at 30 min).
ev_q8h <- make_events(cohort, daily_mg = 12000, tau = 8, label = "4 g q8h")
sim <- rxode2::rxSolve(mod, events = ev_q8h, keep = c("CRCL", "WT", "ALB", "regimen")) |>
  as.data.frame()
stopifnot(nrow(sim) > 0, !all(is.na(sim$Cc)))
```

## Check 2 -- covariate algebra reproduces the printed equations exactly

`rxSolve` returns the realised per-subject `cl` and `vc`. With the random
effects zeroed they must equal the paper's p. 6 equations evaluated at each
subject's covariates. This is deterministic -- no cohort noise -- so it is
asserted tightly, and it is the check that would catch a mis-transcribed
exponent or centring constant.

```{r covariate-algebra}
sim_typ <- rxode2::rxSolve(rxode2::zeroRe(mod), events = ev_q8h,
                           keep = c("CRCL", "WT", "ALB")) |>
  as.data.frame()

chk <- sim_typ |>
  distinct(id, CRCL, WT, ALB, cl, vc) |>
  mutate(
    # Dong 2026 p. 6, with the e^omega^2 eta factor set to 1 (typical subject)
    cl_paper = 6.48 * (CRCL / 46.56)^0.615 * (WT / 70)^1.13,
    vc_paper = 19   * (ALB  / 34.8)^1.21,
    cl_relerr = abs(cl - cl_paper) / cl_paper,
    vc_relerr = abs(vc - vc_paper) / vc_paper
  )

stopifnot(nrow(chk) == n_sub,                 # guard: the check had rows to test
          max(chk$cl_relerr) < 1e-8,
          max(chk$vc_relerr) < 1e-8)

knitr::kable(
  chk |> slice_head(n = 5) |>
    select(id, CRCL, WT, ALB, cl, cl_paper, vc, vc_paper) |>
    dplyr::rename("eGFR" = CRCL, "CL (model)" = cl, "CL (p. 6 equation)" = cl_paper,
                  "V (model)" = vc, "V (p. 6 equation)" = vc_paper),
  digits = 3,
  caption = "First five subjects: model-realised CL and V against the paper's printed equations (max relative error over all 200 subjects < 1e-8)."
)
```

## Check 3 -- steady-state mass balance

At steady state the amount cleared over one dosing interval must equal the dose
delivered in that interval: `CL * AUC(0-tau) = dose`. This is an exact identity
for a linear one-compartment model, so it is asserted tightly. It is worth
running because it fails loudly if the ODE is ever short-circuited -- for
instance by rxode2 auto-solving a `cl`/`vc` pair and discarding the explicit
`d/dt`, or by a bioavailability term zeroing the input.

```{r mass-balance}
mb <- sim_typ |>
  filter(!is.na(Cc)) |>
  group_by(id, cl) |>
  arrange(time, .by_group = TRUE) |>
  summarise(
    auctau = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    .groups = "drop"
  ) |>
  mutate(recovered = cl * auctau / 4000)   # 4 g per q8h interval

# Tolerance covers trapezoidal error on the 0.02 h grid plus the last traces of
# accumulation in the slowest-clearing subjects. It is far tighter than any
# structural break would produce: auto-solving the ODE away or zeroing the
# input moves this to 0 or by whole percent, not by 0.2%.
stopifnot(nrow(mb) == n_sub, max(abs(mb$recovered - 1)) < 2e-3)
cat(sprintf("CL * AUCtau / dose: min %.6f, max %.6f (exact identity = 1)\n",
            min(mb$recovered), max(mb$recovered)))
```

# Replicate published figures

## Figure 1 -- concentration versus time after the last dose

Figure 1 of the source plots all 117 observed concentrations against time since
the last dose, for the clinical 4 g q8h regimen. Observed values span
1.25-376.34 mg/L and the visible cloud falls away from roughly 375 mg/L near the
end of infusion to single digits by 10-12 h. The simulated interval below is the
model's counterpart, with residual error applied so that the spread is
comparable to observed data.

```{r figure-1}
sim |>
  filter(!is.na(Cc)) |>
  mutate(tsld = time - min(time)) |>
  group_by(tsld) |>
  summarise(Q05 = quantile(sim, 0.05), Q50 = quantile(sim, 0.50),
            Q95 = quantile(sim, 0.95), .groups = "drop") |>
  ggplot(aes(tsld, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(linewidth = 0.8) +
  labs(x = "Time after last dose (h)", y = "Piperacillin concentration (mg/L)",
       title = "4 g q8h at steady state: median and 5th-95th percentiles",
       caption = "Counterpart to Figure 1 of Dong 2026 (observed range 1.25-376.34 mg/L).") +
  theme_bw()
```

```{r figure-1-range}
# `sim$sim` is the residual-error-perturbed observation; `sim$Cc` is the
# error-free individual prediction.
dv <- sim$sim[!is.na(sim$sim)]
cat(sprintf("Simulated 4 g q8h steady-state interval: 1st pctile %.2f, 99th pctile %.1f mg/L\n",
            quantile(dv, 0.01), quantile(dv, 0.99)))
# The observed 1.25-376.34 mg/L range pools every sample from every patient and
# every regimen, so it is an envelope rather than a per-percentile target. Assert
# only that the simulated interval lands inside a plausible multiple of it.
stopifnot(quantile(dv, 0.99) < 3 * 376.34, quantile(dv, 0.50) > 1.25)
```

## Figures 5-7 -- probability of target attainment

The paper's Monte Carlo simulations are its main quantitative output. Free
piperacillin is taken as 70% of total (the paper's stated literature
assumption), and three targets are evaluated over one steady-state dosing
interval: 50% *f*T>MIC, 100% *f*T>MIC and 100% *f*T>4xMIC, at the CLSI
breakpoints of 1, 8 and 16 mg/L.

```{r pta-compute}
FU <- 0.70   # unbound fraction, Dong 2026 Methods ("assumed to be 70%")

regimens <- tibble::tribble(
  ~label,           ~daily_mg, ~tau,
  "12 g/day q12h",     12000,    12,
  "12 g/day q8h",      12000,     8,
  "12 g/day q6h",      12000,     6,
  "12 g/day q4h",      12000,     4,
  "20 g/day CII",      20000,    NA,
  "24 g/day CII",      24000,    NA
)

pta_one <- function(label, daily_mg, tau) {
  ev <- make_events(cohort, daily_mg = daily_mg, tau = tau, label = label)
  s  <- rxode2::rxSolve(mod, events = ev) |> as.data.frame()
  s |>
    filter(!is.na(Cc)) |>
    mutate(fC = FU * Cc) |>
    group_by(id) |>
    summarise(
      `50% fT>MIC, MIC 1`      = mean(fC > 1)  >= 0.5,
      `50% fT>MIC, MIC 8`      = mean(fC > 8)  >= 0.5,
      `50% fT>MIC, MIC 16`     = mean(fC > 16) >= 0.5,
      `100% fT>MIC, MIC 1`     = all(fC > 1),
      `100% fT>MIC, MIC 8`     = all(fC > 8),
      `100% fT>MIC, MIC 16`    = all(fC > 16),
      `100% fT>4xMIC, MIC 1`   = all(fC > 4),
      `100% fT>4xMIC, MIC 8`   = all(fC > 32),
      `100% fT>4xMIC, MIC 16`  = all(fC > 64),
      .groups = "drop"
    ) |>
    summarise(across(-id, ~ 100 * mean(.x))) |>
    mutate(regimen = label, .before = 1)
}

pta <- do.call(bind_rows, Map(pta_one, regimens$label,
                              regimens$daily_mg, regimens$tau))
stopifnot(nrow(pta) == nrow(regimens))

knitr::kable(pta, digits = 1,
             caption = "Probability of target attainment (%) by regimen and target. Replicates Figure 5 of Dong 2026.")
```

```{r pta-plot, fig.height = 5}
pta |>
  tidyr::pivot_longer(-regimen, names_to = "target", values_to = "PTA") |>
  tidyr::separate(target, into = c("target", "MIC"), sep = ", ") |>
  mutate(regimen = factor(regimen, levels = regimens$label),
         MIC = factor(MIC, levels = c("MIC 1", "MIC 8", "MIC 16"))) |>
  ggplot(aes(regimen, PTA, fill = MIC)) +
  geom_col(position = "dodge") +
  geom_hline(yintercept = 90, linetype = "dashed") +
  facet_wrap(~target, ncol = 1) +
  coord_cartesian(ylim = c(0, 100)) +
  labs(x = NULL, y = "PTA (%)",
       caption = "Replicates Figure 5 of Dong 2026; dashed line is the 90% PTA criterion.") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Check 4 -- the paper's stated PTA conclusions

The Results section makes a series of explicit claims about which regimens clear
the 90% PTA bar. Each is evaluated below against the simulated cohort. The
`Deviation` column marks claims the model does not reproduce; those are excluded
from the gate and discussed in prose rather than tuned away.

```{r pta-claims}
cell <- function(reg, col) {
  v <- pta[[col]][pta$regimen == reg]
  if (length(v) != 1L) stop("no unique PTA row for '", reg, "' / '", col, "'")
  v
}

claims <- tibble::tribble(
  ~Claim, ~Achieved, ~Pass,
  "50% fT>MIC, MIC 16: q12h fails to reach 90% PTA",
    cell("12 g/day q12h", "50% fT>MIC, MIC 16"),
    cell("12 g/day q12h", "50% fT>MIC, MIC 16") < 90,
  "100% fT>MIC, MIC 1: q6h or more frequent reaches 90% PTA",
    cell("12 g/day q6h", "100% fT>MIC, MIC 1"),
    cell("12 g/day q6h", "100% fT>MIC, MIC 1") >= 90,
  "100% fT>MIC, MIC 8: CII reaches 90% PTA",
    cell("24 g/day CII", "100% fT>MIC, MIC 8"),
    cell("24 g/day CII", "100% fT>MIC, MIC 8") >= 90,
  "100% fT>MIC, MIC 16: CII reaches 90% PTA",
    cell("24 g/day CII", "100% fT>MIC, MIC 16"),
    cell("24 g/day CII", "100% fT>MIC, MIC 16") >= 90,
  "100% fT>4xMIC, MIC 1: q4h or more frequent reaches 90% PTA",
    cell("12 g/day q4h", "100% fT>4xMIC, MIC 1"),
    cell("12 g/day q4h", "100% fT>4xMIC, MIC 1") >= 90,
  "100% fT>4xMIC, MIC 8: CII at >=20 g/day reaches 90% PTA",
    cell("20 g/day CII", "100% fT>4xMIC, MIC 8"),
    cell("20 g/day CII", "100% fT>4xMIC, MIC 8") >= 90,
  "100% fT>4xMIC, MIC 16: only 24 g/day CII approaches 90% PTA; paper reports 78%",
    cell("24 g/day CII", "100% fT>4xMIC, MIC 16"),
    abs(cell("24 g/day CII", "100% fT>4xMIC, MIC 16") - 78) < 15
) |>
  mutate(Deviation = FALSE)

knitr::kable(claims |> mutate(Achieved = round(Achieved, 1)) |>
               dplyr::rename("Achieved PTA (%)" = Achieved),
             caption = "Dong 2026 Results claims evaluated against the packaged model.")

# The headline number is the paper's single most specific PTA figure. Realised
# 79.0 / 81.5 / 84.5 / 85.5 at 16 / 2 / 1 / 4 solver threads against the
# published 78, i.e. a 1.0-7.5 point gap depending on the cohort drawn; the
# 15-point bound sits outside that spread. It can still go red -- a
# mis-transcribed CL or dose moves this by tens of points -- but note it does
# NOT discriminate the omega scale, which Check 1 settles from the source.
stopifnot(nrow(claims) == 7L, all(claims$Pass[!claims$Deviation]))
```

## Figures 6 and 7 -- PTA across renal function and body weight

The paper reports two monotone gradients: PTA falls as eGFR rises (Figure 6) and
falls as total body weight rises (Figure 7), both effects becoming pronounced at
the aggressive 100% *f*T>4xMIC target. Both follow directly from the positive
exponents on CL.

```{r figure-6-7, fig.height = 4}
gradient <- function(varname, values, other) {
  do.call(bind_rows, lapply(seq_along(values), function(i) {
    cov1 <- tibble::tibble(id = 1L, CRCL = other$CRCL, WT = other$WT, ALB = other$ALB)
    cov1[[varname]] <- values[i]
    ev <- make_events(cov1, daily_mg = 24000, tau = NA)
    s <- rxode2::rxSolve(rxode2::zeroRe(mod), events = ev) |> as.data.frame()
    fC <- FU * s$Cc[!is.na(s$Cc)]
    tibble::tibble(level = values[i],
                   `100% fT>MIC (MIC 16)`    = all(fC > 16),
                   `100% fT>4xMIC (MIC 16)`  = all(fC > 64),
                   Css = mean(s$Cc, na.rm = TRUE))
  }))
}
ref <- list(CRCL = 46.56, WT = 70, ALB = 34.8)

g_egfr <- gradient("CRCL", c(20, 40, 60, 90, 130), ref)
g_wt   <- gradient("WT",   c(50, 70, 90, 110, 130), ref)

knitr::kable(
  bind_rows(
    g_egfr |> mutate(Covariate = "eGFR (mL/min/1.73 m^2)", .before = 1),
    g_wt   |> mutate(Covariate = "Total body weight (kg)", .before = 1)
  ) |> dplyr::rename("Level" = level, "Typical Css (mg/L)" = Css),
  digits = 1,
  caption = "Typical-subject steady-state Css and target attainment on 24 g/day CII across renal function (Figure 6) and body weight (Figure 7)."
)

# Both gradients are strictly monotone for a typical subject because the
# exponents on CL are positive; this is deterministic (zeroRe), not a cohort
# statistic, so exact monotonicity is the right assertion here.
stopifnot(nrow(g_egfr) == 5L, nrow(g_wt) == 5L,
          all(diff(g_egfr$Css) < 0), all(diff(g_wt$Css) < 0))
```

The paper states that "achieving the target of 100% *f*T > 4xMIC (MIC=16 mg/L)
became challenging for patients with eGFR exceeding 60 mL/min/1.73 m^2" even on
the highest simulated dose. The typical-subject gradient above reproduces that:
on 24 g/day CII the target holds at low eGFR and is lost as eGFR climbs.

# PKNCA validation

The source paper publishes **no** NCA table -- Cmax, Tmax, AUC and half-life are
never reported -- so there is no published column to compare against and
`ncaComparisonTable()` is not applicable. Instead, PKNCA's steady-state output is
checked against the closed form the model implies, which is an exact identity for
a linear one-compartment system and therefore a strict gate.

```{r pknca}
tau <- 8
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(time = time - min(time)) |>
  dplyr::select(id, time, Cc, regimen)

# Guarantee a time-zero anchor per (id, regimen). For this steady-state window
# the trough at the start of the interval is the correct value, so carry the
# first observed concentration rather than a zero.
sim_nca <- sim_nca |>
  dplyr::group_by(id, regimen) |>
  dplyr::arrange(time, .by_group = TRUE) |>
  dplyr::ungroup()
stopifnot(all(sim_nca$time[sim_nca$time == 0] == 0), sum(sim_nca$time == 0) == n_sub)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(sim_nca), Cc ~ time | regimen + id)

dose_df <- sim_nca |>
  dplyr::distinct(id, regimen) |>
  dplyr::mutate(time = 0, amt = 4000)
dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | regimen + id)

# `cmin` is the interval trough for this profile; `ctrough` is deliberately not
# requested, because PKNCA returns NA for it unless a record sits exactly on the
# interval end and the rebased time grid cannot guarantee that to the bit.
intervals <- data.frame(start = 0, end = tau,
                        cmax = TRUE, tmax = TRUE, cmin = TRUE,
                        auclast = TRUE, cav = TRUE)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

knitr::kable(
  nca_wide |>
    tidyr::pivot_longer(-id, names_to = "Parameter", values_to = "v") |>
    dplyr::group_by(Parameter) |>
    dplyr::summarise(Median = median(v), Q05 = quantile(v, .05),
                     Q95 = quantile(v, .95), .groups = "drop") |>
    dplyr::rename("5th pctile" = Q05, "95th pctile" = Q95),
  digits = 2,
  caption = "PKNCA steady-state parameters over one 8 h interval, 4 g q8h (n = 200)."
)
```

## Check 5 -- PKNCA against the model-implied closed form

```{r pknca-closed-form}
cl_by_id <- sim |> distinct(id, cl)

closed <- nca_wide |>
  inner_join(cl_by_id, by = "id") |>
  mutate(
    auc_closed = 4000 / cl,          # AUC(0-tau) at steady state = dose / CL
    auc_relerr = abs(auclast - auc_closed) / auc_closed,
    cav_relerr = abs(cav - auclast / tau) / (auclast / tau)
  )

stopifnot(nrow(closed) == n_sub)
cat(sprintf("AUC(0-tau) vs dose/CL: median rel. error %.4f%%, max %.4f%%\n",
            100 * median(closed$auc_relerr), 100 * max(closed$auc_relerr)))

# The only difference between the two sides is trapezoidal error on a 0.02 h
# grid, so this is a numerical-accuracy check, not a cohort statistic: a tight
# absolute bound is correct and should be kept.
stopifnot(max(closed$auc_relerr) < 0.005, max(closed$cav_relerr) < 1e-6)
```

The agreement is trapezoidal error only -- the two sides use the same drawn
parameters -- which is why the bound is tight here while the PTA bound above is
not.

# Assumptions and deviations

- **The final-model equations are not in the PDF's text layer.** Both equations
  on page 6 are vector graphics and are dropped by every text extraction of this
  paper. They were read from a rendered image of the page. They are the sole
  source for the three centring constants (46.56, 70, 34.8) and for the
  `e^0.114` / `e^0.069` factors that settle the omega scale. Each of the three
  constants was independently confirmed to equal the corresponding Table 1
  cohort median.
- **Omega scale.** Table 3's `IIV-CL (%)` / `IIV-V (%)` column is
  `100 * omega` (a log-scale SD), not a lognormal CV. This is settled by the
  printed equation factors, as shown in Check 1, and is not an interpretation
  choice.
- **Covariate distributions are reconstructed, not published.** The paper
  simulated from "datasets consistent with the characteristics of the original
  dataset" without publishing that joint distribution. Each covariate here is
  drawn lognormally from its Table 1 median and IQR, and the three are drawn
  **independently** -- in reality eGFR, weight and albumin are correlated in ICU
  patients. This is the main reason the reproduced PTA values sit a few points
  above the published ones rather than on top of them.
- **The albumin-on-V exponent is positive, which is the opposite of the
  direction the paper's own Discussion argues for.** The fitted exponent 1.21
  makes a *lower* albumin give a *smaller* V, while the Discussion reasons that
  hypoalbuminaemia should *increase* V through reduced binding and expanded
  extracellular fluid. The fitted value is what is encoded, because it is what
  the model was estimated with and it is consistent between Table 3 and the
  printed equation; but it is the least precisely estimated parameter in the
  model (RSE 33.4%, bootstrap 5th-95th 0.53-2.06) and the paper does not
  reconcile the discrepancy. A user should treat the direction of this effect as
  weakly supported.
- **Weight enters CL with an estimated exponent of 1.13, not a fixed allometric
  0.75**, and does not scale V at all. The cohort IQR is narrow (65-75 kg) while
  the paper's own Figure 7 extrapolates past 110 kg, so the term does
  substantial work outside the range it was fitted in.
- **Infusion duration.** The clinical infusion duration is not reported. All
  intermittent regimens here use the 30 min the paper's own simulations
  standardise on; the single intravenous-push patient (2.38%) is not modelled
  separately.
- **Unbound fraction 0.70** is a literature assumption the paper adopts for its
  target-attainment work, not a fitted parameter. It is used in this vignette's
  PTA calculations only and is deliberately *not* encoded in the model file. The
  assay measured total, not free, piperacillin -- the paper lists this as a
  limitation.
- **Renal replacement therapy.** Patients on CRRT were excluded, so neither the
  model nor these simulations say anything about extracorporeal clearance.
- **Tazobactam is not modelled.** The study measured piperacillin only.
