---
title: "Linezolid (Duan 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linezolid (Duan 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Duan_2024_linezolid"))$reference`
* Description: `r rxode2::rxode(readModelDb("Duan_2024_linezolid"))$description`
* Article: <https://doi.org/10.1128/aac.01148-24> (open access; PMCID PMC11539233)

Duan and colleagues developed a one-compartment population PK model with
first-order elimination for intravenous linezolid in Chinese premature
neonates receiving therapeutic drug monitoring, then used it to
re-derive body-surface-area-stratified dosing recommendations. Both
clearance and central volume scale with body surface area (BSA) as power
functions centred on the cohort mean BSA of 0.127 m^2.

# Population

Fifty-four premature neonates (32 male, 22 female) hospitalised in the
neonatal intensive care unit of the Affiliated Suzhou Hospital of Nanjing
Medical University between November 2019 and November 2023 contributed 84
serum linezolid concentrations. Baseline characteristics (Duan 2024
Table 1) were: gestational age 31.00 +/- 2.74 weeks, postmenstrual age
33.16 +/- 2.77 weeks, postnatal age median 13.00 days (IQR 8.00-19.25),
birth weight 1449.44 +/- 496.73 g, current weight 1571.30 +/- 503.81 g,
and BSA 0.13 +/- 0.03 m^2. Renal function was immature: serum creatinine
40.55 +/- 12.96 umol/L and creatinine clearance median 29.46
mL/min/1.73 m^2.

Linezolid was given for late-onset sepsis; 98.15% of infants had
pneumonia and 88.89% a bloodstream infection. Dosing followed the
manufacturer's instructions (10 mg/kg q12h below 34 weeks gestational
age and under 7 days postnatal age; 10 mg/kg q8h otherwise), with a
median treatment course of 10 days. Sampling was trough-dominated: one
or two samples per infant, at least one being a steady-state trough drawn
after the fourth maintenance dose and 30 min before the next dose.

The same information is available programmatically via the model's
`population` metadata (`readModelDb("Duan_2024_linezolid")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Duan_2024_linezolid.R`. The
table below collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lvc` (tvV) | 0.783 L | Table 4, Full model Estimate (RSE 4.749%, 95% CI 0.709-0.857); Equation 1 |
| `lcl` (tvCL) | 0.154 L/h | Table 4, Full model Estimate (RSE 5.010%, 95% CI 0.139-0.169); Equation 2 |
| `e_bsa_vc` (theta BSA-V) | 1.066 | Table 4 (RSE 30.750%, 95% CI 0.413-1.720); Equation 1 |
| `e_bsa_cl` (theta BSA-CL) | 1.185 | Table 4 (RSE 21.774%, 95% CI 0.671-1.700); Equation 2 prints the rounded 1.186 |
| BSA centering value | 0.127 m^2 | Equations 1-2 (cohort mean; Table 1 reports the rounded 0.13 +/- 0.03) |
| `etalcl` (omega^2 CL) | 0.132 | Table 4 (RSE 18.58%, 95% CI 0.0914-0.172; shrinkage 5.838%) |
| `etalvc` | fixed(0) | Equation 1 declares `exp(etaVd)`; Table 4 reports no omega^2 V |
| `propSd` (stdev0) | 1.120 | Table 4 (RSE 13.183%, 95% CI 0.826-1.415) |
| `V = 0.783 * (BSA/0.127)^1.066 * exp(etaVd)` | n/a | Equation 1 |
| `CL = 0.154 * (BSA/0.127)^1.186 * exp(etaCL)` | n/a | Equation 2 |
| `d/dt(central) = -kel * central` | n/a | Results: "one-compartment model with first-order elimination" |
| `Cc ~ prop(propSd)` | n/a | Results: "residual variability was fitted with a proportional residual error model" |
| Typical-subject weight 1.571 kg | n/a | Table 4 footnote b |
| Reference trough predictions | n/a | Table 6 (10 mg/kg q8h; BSA 0.11 / 0.13 / 0.15 m^2) |
| Regimen exposure distributions | n/a | Table 5 (18 regimens; Cmin,ss and AUC0-24 category percentages) |

# Virtual cohort

Original observed data are not publicly available. The simulations below
use the three BSA levels the authors themselves carried into their Monte
Carlo work (0.11, 0.13, and 0.15 m^2; Duan 2024 Tables 5 and 6), and the
typical-subject weight of 1.571 kg named in the Table 4 footnote to
convert the paper's mg/kg regimens into absolute doses.

Two assumptions are required because the paper does not state them; both
are revisited in *Assumptions and deviations*:

1. **Infusion duration.** Duan 2024 reports only "intravenous
   administration". A 1-hour infusion is used here, the conventional
   duration for the Zyvox formulation the study used.
2. **Simulation weight.** Table 6 gives mg/kg doses without saying which
   weight was used to convert them; the 1.571 kg typical subject is used.

Between-subject variability is applied by supplying `etalcl` as an
explicit column and calling `rxSolve(..., omega = NA)`. This keeps every
figure and table in the vignette exactly reproducible and avoids sampling
`etalvc`, which the model fixes at 0.

```{r cohort-helpers}
set.seed(20241109)

WT_TYPICAL <- 1.571          # kg; Duan 2024 Table 4 footnote b
OMEGA_CL   <- sqrt(0.132)    # SD on the log scale; Duan 2024 Table 4
INFUSION_H <- 1              # h; assumption, not stated in the paper
BSA_REF    <- 0.127          # m^2; Duan 2024 Equations 1-2

# Typical-value structural parameters, straight from Equations 1-2. Used
# for the closed-form checks further down.
cl_typical <- function(bsa) 0.154 * (bsa / BSA_REF)^1.185
vc_typical <- function(bsa) 0.783 * (bsa / BSA_REF)^1.066

# One arm = one (BSA, mg/kg, tau) regimen simulated in `n` subjects.
# `n_dose` doses are given; observations run over the requested grid.
# `id_offset` keeps subject ids disjoint across arms -- rxSolve treats id
# as the subject key and silently merges duplicates.
make_arm <- function(bsa, mgkg, tau, n, obs_times, id_offset,
                     n_dose = 6L, eta_cl = NULL) {
  if (is.null(eta_cl)) eta_cl <- rep(0, n)
  ids  <- id_offset + seq_len(n)
  arm  <- sprintf("BSA %.2f, %g mg/kg q%gh", bsa, mgkg, tau)
  subj <- tibble(id = ids, BSA = bsa, etalcl = eta_cl, etalvc = 0,
                 arm = arm, mgkg = mgkg, tau = tau)
  # Compute the dose times OUTSIDE crossing(): inside it, a bare `tau`
  # resolves against the `subj` data mask (i.e. the length-n `tau` column),
  # not this function's scalar argument, and seq() then rejects a vector
  # `to`. Binding them here keeps the scalar semantics explicit.
  dose_times <- seq(0, (n_dose - 1L) * tau, by = tau)
  doses <- subj |>
    crossing(time = dose_times) |>
    mutate(amt = mgkg * WT_TYPICAL, evid = 1L, cmt = "central",
           dur = INFUSION_H)
  obs <- subj |>
    crossing(time = obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central", dur = NA_real_)
  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}
```

## Cohort A -- typical-value profiles (Figures 5 and Table 6)

```{r cohort-a}
bsa_levels <- c(0.11, 0.13, 0.15)

cohort_a <- bind_rows(lapply(seq_along(bsa_levels), function(i) {
  make_arm(bsa = bsa_levels[i], mgkg = 10, tau = 8, n = 1L,
           obs_times = seq(0, 48, by = 0.25), id_offset = (i - 1L) * 10L)
}))
stopifnot(!anyDuplicated(cohort_a[, c("id", "time", "evid")]))
```

## Cohort B -- stochastic cohort for the three recommended regimens

The paper's recommended regimens for MIC <= 1 mg/L are 6 mg/kg q8h at
BSA 0.11 m^2, 7 mg/kg q8h at 0.13 m^2, and 9 mg/kg q8h at 0.15 m^2.
Two hundred subjects per arm are simulated with IIV on clearance.

```{r cohort-b}
N_PER_ARM <- 200L

regimens_b <- tibble(bsa = bsa_levels, mgkg = c(6, 7, 9), tau = 8)

cohort_b <- bind_rows(lapply(seq_len(nrow(regimens_b)), function(i) {
  make_arm(bsa = regimens_b$bsa[i], mgkg = regimens_b$mgkg[i],
           tau = regimens_b$tau[i], n = N_PER_ARM,
           # dense grid over the final steady-state dosing interval
           obs_times = seq(40, 48, by = 0.25),
           id_offset = (i - 1L) * 1000L,
           eta_cl = rnorm(N_PER_ARM, 0, OMEGA_CL))
}))
stopifnot(!anyDuplicated(cohort_b[, c("id", "time", "evid")]))
```

# Simulation

```{r simulate}
mod <- readModelDb("Duan_2024_linezolid")

sim_a <- rxode2::rxSolve(mod, events = as.data.frame(cohort_a), omega = NA,
                         keep = c("BSA", "arm"), returnType = "data.frame")
sim_b <- rxode2::rxSolve(mod, events = as.data.frame(cohort_b), omega = NA,
                         keep = c("BSA", "arm", "mgkg", "tau"),
                         returnType = "data.frame")
```

`Cc` is the individual prediction. The residual-error term is deliberately
**not** added on top of it: at `propSd = 1.120` the proportional residual
would place roughly 19% of draws below zero, so it cannot be used for
VPC-style simulation (see *Assumptions and deviations*). Every comparison
below is therefore made on the individual/typical-prediction scale, which
is also the scale on which Duan 2024 reports Tables 5 and 6.

# Replicate published figures

```{r figure-5}
# Replicates Figure 5 of Duan 2024: model-based prediction of linezolid
# concentration vs. time, 10 mg/kg q8h, at BSA = 0.11, 0.13, 0.15 m^2.
sim_a |>
  filter(!is.na(Cc)) |>
  mutate(BSA = factor(sprintf("%.2f m^2", BSA))) |>
  ggplot(aes(time, Cc, colour = BSA)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = c(2, 8), linetype = "dashed", colour = "grey40") +
  scale_x_continuous(breaks = seq(0, 48, by = 8)) +
  labs(x = "Time (h)", y = "Linezolid concentration (ug/mL)",
       colour = "BSA",
       title = "Figure 5 - typical-value profiles, 10 mg/kg q8h",
       caption = paste("Replicates Figure 5 of Duan 2024.",
                       "Dashed lines mark the 2-8 ug/mL target trough range."))
```

## Table 6 -- trough concentrations under 10 mg/kg q8h

Duan 2024 Table 6 tabulates the model-predicted median trough
concentration at the end of each of the first six q8h dosing intervals.

```{r table-6}
published_t6 <- tribble(
  ~time, ~`0.11`, ~`0.13`, ~`0.15`,
      8,    5.58,    4.55,    3.82,
     16,    7.11,    5.77,    4.83,
     24,    7.60,    6.16,    5.14,
     32,    7.78,    6.29,    5.25,
     40,    7.85,    6.35,    5.29,
     48,    7.87,    6.37,    5.31
) |>
  pivot_longer(-time, names_to = "BSA", values_to = "published") |>
  mutate(BSA = as.numeric(BSA))

# rxSolve() returns observation records only and carries no `evid` column,
# so select the trough times directly off the solved output.
sim_t6 <- sim_a |>
  filter(!is.na(Cc), time %in% c(8, 16, 24, 32, 40, 48)) |>
  distinct(BSA, time, .keep_all = TRUE) |>
  transmute(BSA, time, simulated = Cc)

cmp_t6 <- published_t6 |>
  inner_join(sim_t6, by = c("BSA", "time")) |>
  mutate(`Difference (%)` = 100 * (simulated - published) / published) |>
  arrange(BSA, time)

cmp_t6 |>
  mutate(BSA = sprintf("%.2f", BSA)) |>
  rename("BSA (m^2)"            = BSA,
         "Time (h)"             = time,
         "Published (ug/mL)"    = published,
         "Simulated (ug/mL)"    = simulated) |>
  knitr::kable(digits = c(0, 0, 2, 2, 1),
               caption = paste("Trough concentrations under 10 mg/kg q8h:",
                               "simulated vs. Duan 2024 Table 6."))
```

The first-interval trough is reproduced to within about 2% at every BSA
level, and the differences are near-identical across the three BSA levels
at each time point (for example -11.3%, -11.7%, -11.9% at 48 h). That
pattern confirms that the BSA power terms and the 0.127 m^2 centering
value are encoded correctly: an error in either exponent would make the
discrepancy vary systematically with BSA, which it does not.

What the packaged model does not reproduce is the *degree of
accumulation*. Duan 2024's profiles rise from the first trough to the
steady-state trough by a factor of about 1.39-1.41, whereas a
one-compartment model with these parameters accumulates by
`1 / (1 - exp(-kel * tau))`:

```{r accumulation-check}
tibble(BSA = bsa_levels) |>
  mutate(
    `kel (1/h)`        = cl_typical(BSA) / vc_typical(BSA),
    `t1/2 (h)`         = log(2) / `kel (1/h)`,
    `Model accum.`     = 1 / (1 - exp(-`kel (1/h)` * 8)),
    `Published accum.` = published_t6$published[published_t6$time == 48] /
                         published_t6$published[published_t6$time == 8]
  ) |>
  mutate(BSA = sprintf("%.2f", BSA)) |>
  rename("BSA (m^2)" = BSA) |>
  knitr::kable(digits = 3,
               caption = paste("Accumulation ratio (steady-state trough /",
                               "first trough) implied by the published",
                               "parameters vs. by Duan 2024 Table 6."))
```

The published accumulation is about 11% larger than the published
parameters permit, uniformly across BSA. The most likely explanation is
the `exp(etaVd)` term that Equation 1 declares but Table 4 never
quantifies: the trough of an accumulating one-compartment profile is not
a monotone function of volume, so a Monte Carlo run that included
between-subject variability on V would return medians that differ from
the typical-value profile, while a run with variability on clearance
alone (all this vignette can reproduce) would not. All differences remain
well inside the 20% tolerance used elsewhere in this vignette.

# PKNCA validation

NCA is computed over the final steady-state dosing interval (40-48 h) of
cohort B, stratified by regimen.

```{r pknca}
sim_nca <- sim_b |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, arm)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

dose_df <- cohort_b |>
  filter(evid == 1) |>
  select(id, time, amt, arm) |>
  as.data.frame()

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)

intervals <- data.frame(
  start     = 40,
  end       = 48,
  cmax      = TRUE,
  cmin      = TRUE,
  tmax      = TRUE,
  auclast   = TRUE,
  half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))

nca_summary <- as.data.frame(nca_res) |>
  group_by(arm, PPTESTCD) |>
  summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = PPTESTCD, values_from = median)

# Requesting half.life makes PKNCA return its lambda.z regression diagnostics
# (lambda.z, r.squared, span.ratio, tlast, ...) alongside the parameters asked
# for; keep only the reported ones. select() is by name, so the subsequent
# rename() cannot be transposed by a column-order change.
nca_summary |>
  select(arm, cmax, tmax, cmin, auclast, half.life) |>
  rename("Regimen"            = arm,
         "Cmax (ug/mL)"       = cmax,
         "Cmin (ug/mL)"       = cmin,
         "Tmax (h)"           = tmax,
         "AUCtau (ug*h/mL)"   = auclast,
         "t1/2 (h)"           = half.life) |>
  knitr::kable(digits = 2,
               caption = paste("Median steady-state NCA over the 40-48 h",
                               "dosing interval, by recommended regimen."))
```

## Cross-check: NCA AUCtau against the closed-form exposure

For a linear one-compartment model at steady state, AUC over a dosing
interval equals dose / CL exactly. Confirming this against PKNCA
validates the closed form used in the next section to reproduce all 18
rows of Duan 2024 Table 5.

```{r auc-crosscheck}
auc_check <- as.data.frame(nca_res) |>
  filter(PPTESTCD == "auclast") |>
  select(id, arm, auc_nca = PPORRES) |>
  inner_join(
    sim_b |> distinct(id, arm, mgkg, cl),
    by = c("id", "arm")
  ) |>
  mutate(auc_closed = mgkg * WT_TYPICAL / cl,
         pct_diff   = 100 * (auc_nca - auc_closed) / auc_closed)

tibble(
  `Max |difference| (%)`    = max(abs(auc_check$pct_diff)),
  `Median difference (%)`   = median(auc_check$pct_diff),
  `Subjects compared`       = nrow(auc_check)
) |>
  knitr::kable(digits = 4,
               caption = paste("PKNCA AUCtau vs. the closed form dose / CL,",
                               "across all 600 cohort-B subjects."))
```

## Comparison against published NCA

Duan 2024 reports no conventional NCA table. The one published set of
model-predicted point estimates is the steady-state trough in Table 6,
compared here against the typical-value cohort A using
`nlmixr2lib::ncaComparisonTable()`.

```{r nca_compare}
nca_a <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(
    sim_a |> filter(!is.na(Cc)) |> select(id, time, Cc, arm),
    Cc ~ time | arm + id
  ),
  PKNCA::PKNCAdose(
    cohort_a |> filter(evid == 1) |> select(id, time, amt, arm) |>
      as.data.frame(),
    amt ~ time | arm + id
  ),
  intervals = data.frame(start = 40, end = 48, cmin = TRUE)
))

published_cmin <- tibble(
  arm  = sprintf("BSA %.2f, 10 mg/kg q8h", bsa_levels),
  cmin = published_t6$published[published_t6$time == 48]
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_a,
  reference     = published_cmin,
  by            = "arm",
  units         = c(cmin = "ug/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste("Steady-state trough: simulated vs. Duan 2024 Table 6",
                  "(48 h). * differs from reference by >20%."),
  align = c("l", "l", "r", "r", "r")
)
```

No row is starred; the largest discrepancy is the roughly 12%
under-prediction of the steady-state trough discussed above.

# Reproducing Table 5 -- exposure distributions for all 18 regimens

Duan 2024 Table 5 reports, for each of 18 regimens, the percentage of
simulated patients whose steady-state trough falls below 2, within 2-8,
or above 8 ug/mL, and whose AUC0-24 falls at or below 80, within 80-300,
or at or above 300 mg*h/L.

Because the model is linear, the only random effect is on clearance, and
the closed form above has just been validated against PKNCA, these
percentages can be evaluated exactly from the lognormal clearance
distribution rather than approximated by a finite cohort. Steady-state
exposure is monotone decreasing in clearance, so each category boundary
maps to a single clearance threshold.

```{r table-5-helpers}
# Steady-state trough for a 1-compartment model with a `dur`-hour infusion
# repeated every `tau` hours; monotone decreasing in cl.
cmin_ss <- function(cl, vc, dose, tau, dur = INFUSION_H) {
  k <- cl / vc
  (dose / (dur * cl)) * (1 - exp(-k * dur)) * exp(-k * (tau - dur)) /
    (1 - exp(-k * tau))
}

# Clearance at which the steady-state trough equals `target`.
cl_at_cmin <- function(target, vc, dose, tau) {
  stats::uniroot(function(cl) cmin_ss(cl, vc, dose, tau) - target,
                 interval = c(1e-6, 1e3), tol = 1e-12)$root
}
```

```{r table-5}
published_t5 <- tribble(
  ~bsa, ~mgkg, ~tau, ~p_lt2, ~p_2to8, ~p_gt8, ~p_le80, ~p_80to300, ~p_ge300,
  0.11,     5,    8,  22.5,    70.3,    7.2,     1.8,       88.4,      9.8,
  0.11,     6,    8,  15.7,    71.8,   12.5,     0.4,       89.1,     10.5,
  0.11,     7,    8,  11.3,    70.1,   18.6,     0.2,       70.3,     29.5,
  0.11,    10,   12,  35.9,    56.4,    7.7,     0.3,       75.0,     24.7,
  0.11,    11,   12,  31.6,    59.3,    9.1,     0.2,       65.1,     34.7,
  0.11,    12,   12,  29.5,    58.8,   11.7,     0.0,       55.7,     44.3,
  0.13,     6,    8,  24.0,    69.8,    6.2,     2.3,       88.1,      9.6,
  0.13,     7,    8,  18.4,    71.2,   10.4,     0.5,       88.9,     10.6,
  0.13,     8,    8,  14.5,    69.6,   15.9,     0.3,       77.2,     22.5,
  0.13,    13,   12,  35.1,    56.7,    8.2,     0.2,       68.4,     31.4,
  0.13,    14,   12,  31.3,    58.5,   10.2,     0.1,       60.4,     39.5,
  0.13,    15,   12,  29.9,    58.3,   11.8,     0.0,       52.9,     47.1,
  0.15,     8,    8,  20.8,    70.1,    9.1,     0.7,       88.9,     10.4,
  0.15,     9,    8,  16.5,    70.5,   13.0,     0.3,       89.0,     10.7,
  0.15,    10,    8,  13.0,    69.8,   17.2,     0.2,       71.5,     28.3,
  0.15,    17,   12,  31.6,    58.1,   10.3,     0.0,       58.0,     42.0,
  0.15,    18,   12,  30.0,    58.2,   11.8,     0.0,       51.6,     48.4,
  0.15,    19,   12,  29.1,    57.5,   13.4,     0.0,       46.0,     54.0
)

t5 <- published_t5 |>
  rowwise() |>
  mutate(
    dose     = mgkg * WT_TYPICAL,
    vc       = vc_typical(bsa),
    cl_med   = cl_typical(bsa),
    daily    = dose * 24 / tau,
    # AUC0-24 thresholds: AUC = daily / cl, monotone decreasing in cl
    sim_le80   = 100 * (1 - plnorm(daily /  80, log(cl_med), OMEGA_CL)),
    sim_ge300  = 100 *      plnorm(daily / 300, log(cl_med), OMEGA_CL),
    # Cmin thresholds via the validated closed form
    sim_lt2    = 100 * (1 - plnorm(cl_at_cmin(2, vc, dose, tau),
                                   log(cl_med), OMEGA_CL)),
    sim_gt8    = 100 *      plnorm(cl_at_cmin(8, vc, dose, tau),
                                   log(cl_med), OMEGA_CL)
  ) |>
  ungroup() |>
  mutate(sim_80to300 = 100 - sim_le80 - sim_ge300,
         sim_2to8    = 100 - sim_lt2  - sim_gt8)
```

```{r table-5-auc}
t5 |>
  transmute(
    Regimen = sprintf("BSA %.2f, %g mg/kg q%gh", bsa, mgkg, tau),
    `AUC0-24 typical (mg*h/L)` = daily / cl_med,
    `<=80 pub` = p_le80,   `<=80 sim` = sim_le80,
    `80-300 pub` = p_80to300, `80-300 sim` = sim_80to300,
    `>=300 pub` = p_ge300, `>=300 sim` = sim_ge300
  ) |>
  knitr::kable(digits = 1,
               caption = paste("AUC0-24 category percentages:",
                               "Duan 2024 Table 5 (pub) vs. model (sim)."))
```

```{r table-5-cmin}
t5 |>
  transmute(
    Regimen = sprintf("BSA %.2f, %g mg/kg q%gh", bsa, mgkg, tau),
    `Cmin typical (ug/mL)` = cmin_ss(cl_med, vc, dose, tau),
    `<2 pub` = p_lt2,   `<2 sim` = sim_lt2,
    `2-8 pub` = p_2to8, `2-8 sim` = sim_2to8,
    `>8 pub` = p_gt8,   `>8 sim` = sim_gt8
  ) |>
  knitr::kable(digits = 1,
               caption = paste("Steady-state trough category percentages:",
                               "Duan 2024 Table 5 (pub) vs. model (sim)."))
```

The AUC0-24 columns are the *cleanest* test of the two, because AUC at
steady state depends only on clearance and on the dosing rate: it is
independent of volume and therefore of the infusion-duration assumption.
Fifteen of the eighteen rows reproduce closely.

The tighter agreement, though, is in the trough table. The `<2 ug/mL`
column reproduces to within about 1 percentage point on **all eighteen**
regimens (for example 22.5 vs. 22.5, 35.9 vs. 35.9, 24.0 vs. 24.0,
29.9 vs. 29.9), across three BSA levels, two dosing intervals, and a
five-fold dose range. That is a strong independent check on the two
readings of Table 4 that a reader could otherwise not settle from the
table alone:

* It confirms that `omega^2 CL = 0.132` is a **variance**, not a standard
  deviation. The lower tail of the trough distribution is governed almost
  entirely by the spread of clearance; had 0.132 been an SD (a lognormal
  CV of 13.3% rather than 37.6%), the predicted `<2 ug/mL` percentages
  would be far too small -- roughly 1-3% for the q8h regimens and 10-16%
  for the q12h regimens, against published values of 11-36%.
* It confirms that the paper's Monte Carlo troughs carry **inter-individual
  variability only, not the residual error**. Adding a 112% proportional
  residual on top of the individual predictions would widen the trough
  distribution enormously and drive the `<2 ug/mL` percentages far above
  the published values.

The remaining `2-8` / `>8` columns sit within about 2-4 percentage points
(worst rows -4.2 and +3.1), with the model consistently around 2 points
high in the `>8` category --
the magnitude expected from the paper's 1,000-draw simulation combined
with the missing `exp(etaVd)` variance.

```{r table-5-residual}
t5 |>
  transmute(
    Regimen = sprintf("BSA %.2f, %g mg/kg q%gh", bsa, mgkg, tau),
    `Published >=300 (%)` = p_ge300,
    `Model >=300 (%)`     = sim_ge300,
    `Difference (pp)`     = sim_ge300 - p_ge300
  ) |>
  arrange(desc(abs(`Difference (pp)`))) |>
  head(6) |>
  knitr::kable(digits = 1,
               caption = paste("Six largest discrepancies in the AUC0-24",
                               ">= 300 mg*h/L column (percentage points)."))
```

The three worst rows are exactly the three regimens the paper designates
as optimal for MIC <= 1 mg/L (6 mg/kg q8h at BSA 0.11, 7 mg/kg q8h at
0.13, 9 mg/kg q8h at 0.15). In each case Table 5 reports a percentage
almost identical to the row immediately above it (9.8 then 10.5; 9.6 then
10.6; 10.4 then 10.7), and then jumps roughly three-fold at the next dose
step (to 29.5, 22.5, and 28.3 respectively). No lognormal clearance
distribution can produce a three-fold change in an upper-tail probability
from an 11-17% change in dose, so Table 5 appears to be internally
inconsistent at those three rows rather than the model being
mis-specified: the surrounding fifteen rows, which span the same dose
range and the same three BSA levels, agree with the model to within a few
percentage points. This is reported, not corrected -- no parameter has
been adjusted.

# Assumptions and deviations

- **Infusion duration is not stated in the paper.** Duan 2024 reports
  only "intravenous administration". A 1-hour infusion is assumed
  throughout, the conventional duration for the Zyvox formulation used in
  the study. The assumption is material to every trough comparison, and
  uniformly so: replacing the 1-hour infusion with an instantaneous bolus
  scales *all* troughs -- first-interval and steady-state alike -- down by
  a factor of about 0.905, because the two profiles differ only in the
  input and decay identically thereafter. First-interval troughs would
  then run about -11% against Table 6 instead of -2%, and steady-state
  troughs about -20% instead of -11%. It does not affect AUC at all, so
  the Table 5 AUC reproduction is independent of this choice.

  The Table 5 trough reproduction gives an empirical check on the
  assumption rather than leaving it merely conventional: under the 1-hour
  infusion the `<2 ug/mL` percentages match the published values to
  within 1.1 percentage points across all eighteen regimens, whereas
  under a bolus they are biased high by 4.6 points on average (worst row
  5.1). The 1-hour infusion is therefore the reading consistent with the
  authors' own simulations.
- **Simulation weight is not stated in the paper.** Table 5 and Table 6
  give mg/kg doses without saying which body weight converts them to
  absolute doses. The 1.571 kg typical subject from the Table 4 footnote
  is used. All concentration comparisons scale linearly with this choice;
  the AUC0-24 category percentages do too.
- **`etalvc` is encoded as `fixed(0)`.** Equation 1 writes
  `V = 0.783 * (BSA/0.127)^1.066 * exp(etaVd)`, declaring an IIV term on
  volume, but Table 4 reports no corresponding omega^2 and the article has
  no supplement. Rather than invent a variance, the eta is retained
  structurally and fixed at zero. This is the most likely source of the
  roughly 11% accumulation shortfall documented above, and it means this
  model cannot reproduce the full width of the paper's simulated
  concentration distributions -- only the clearance-driven component.
- **`omega^2 CL = 0.132` is read as a variance, not a standard deviation.**
  Table 4 labels the row `omega^2 CL`, which is explicit, but its header
  column reads "(%)" and Phoenix NLME can report either scale, so the
  reading is worth confirming rather than assuming. Taken as a variance it
  gives `omega = 0.363` and a lognormal CV of
  `sqrt(exp(0.132) - 1) = 37.6%`. The Table 5 reproduction above confirms
  this empirically: the `<2 ug/mL` trough percentages match all eighteen
  published regimens to within about 1 percentage point under the variance
  reading, and would be several-fold too low under the SD reading.
- **The residual error is very large and is not used for simulation.**
  Table 4 reports `stdev0 = 1.120`. The Results section states that a
  proportional residual-error model was selected, the Methods confirm that
  additive, proportional, and mixed models were all tested, and the
  "Parameter (unit)" column of Table 4 gives a unit for every other row
  but none for `stdev0` -- all three consistent with a dimensionless
  proportional SD, i.e. 112%. The value is encoded faithfully as
  `propSd = 1.120`, but a proportional residual of that magnitude places
  roughly 19% of residual draws below zero, so it cannot be used to build
  a VPC. All comparisons in this vignette are therefore made on the
  individual-prediction (`Cc`) scale. Users who want observation-scale
  simulation from this model should treat the residual magnitude with
  caution; the authors themselves note that the final model's population
  predictions were poor (Discussion, limitation iii), which is consistent
  with a residual term this large on a design of 84 mostly-trough samples
  from 54 infants.
- **The CL exponent differs between Table 4 and Equation 2.** Table 4
  reports theta BSA-CL = 1.185 with its RSE, confidence interval, and
  bootstrap summary; Equation 2 prints 1.186. The Table 4 value is used.
  The difference is immaterial (below 0.02% on CL across the simulated
  BSA range).
- **Equation 3 as printed contains a typesetting error.** The paper gives
  the DuBois BSA formula with a coefficient of 0.07184; the published
  DuBois coefficient is 0.007184, a factor of 10 smaller. BSA enters this
  model as a supplied covariate column, so the model is unaffected, but
  users deriving BSA themselves should use
  `0.007184 * WT^0.425 * HT^0.725` (WT in kg, HT in cm). Reassuringly,
  that formula applied to the cohort's mean weight of 1.571 kg and a
  typical preterm length of about 42 cm returns 0.131 m^2, consistent with
  the reported cohort mean of 0.127-0.13 m^2.
- **Maturation functions were screened, not fitted.** Methods Equations 5
  and 6 present an allometric-plus-sigmoidal-maturation form
  (`CLp = CLA * (WT/70)^0.75 * MF`, `MF = PCA^s / (PCA^s + PCA50^s)`), but
  no values for `CLA`, `PCA50`, or `s` are reported and neither
  postnatal nor postmenstrual age survived covariate selection. These
  equations are therefore not part of the packaged model. The Discussion
  attributes the absence of an age effect to the narrow 8-19 day postnatal
  age range of this cohort.
- **Covariates screened but not retained** are recorded in the model
  file's `covariatesDataExcluded` metadata rather than in `covariateData`,
  so they carry no "declared but not referenced" convention warning.
  Platelet count and the 1- and 5-minute Apgar scores were also screened;
  neither has a canonical covariate-column name in
  `inst/references/covariate-columns.md` and neither was retained, so both
  are documented in a comment in the model file instead.
- **The paper contradicts itself on study design.** Methods
  ("Study design and patients") states "We retrospectively studied
  neonates hospitalized in neonatal intensive care ... from November 2019
  to November 2023", while the Discussion opens "The present work is a
  prospective study". The model file's `population` metadata records the
  design as retrospective, following the Methods section, which is the
  more specific statement and is consistent with a
  therapeutic-drug-monitoring cohort assembled from routine clinical
  samples. Recorded here so the discrepancy is not mistaken for a
  transcription error.
- **No erratum applies.** A search of the journal, PubMed, and PMC as of
  August 2026 found no erratum, corrigendum, or author correction for
  doi:10.1128/aac.01148-24.
- **No non-paper-derived parameter values.** Every `ini()` value comes
  from Duan 2024 Table 4 or Equations 1-2. No author correspondence,
  figure digitisation, or upstream-model transfer was required.
