---
title: "Monoclonal antibody TMDD model instability (Duffull 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Monoclonal antibody TMDD model instability (Duffull 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Duffull 2025 is a tutorial on diagnosing and resolving population-model
instability. Its **case example 1** works through a two-compartment
target-mediated drug disposition (TMDD) model for an unnamed monoclonal
antibody (mAb), shows that the quasi-steady-state binding constant is not
deterministically identifiable under a realistic clinical sampling design, and
resolves the instability by simplifying the model in the target-saturated
limit. Both the unstable full model and the stable simplification are packaged
here, because the pair is what makes the tutorial's argument reproducible.

* Citation: `r rxode2::rxode(readModelDb("Duffull_2025_mab_tmdd_qss"))$reference`
* Article: <https://doi.org/10.1002/psp4.70049>

**Full model** (`Duffull_2025_mab_tmdd_qss`) --
`r rxode2::rxode(readModelDb("Duffull_2025_mab_tmdd_qss"))$description`

**Simplified model** (`Duffull_2025_mab_tmdd_simplified`) --
`r rxode2::rxode(readModelDb("Duffull_2025_mab_tmdd_simplified"))$description`

## What this paper is, and what is not packaged

This is a methods tutorial, not a drug-development report, and only part of it
yields a packageable model. For the reviewer's benefit:

* **Case example 1 (TMDD, packaged as the two models above).** Equations 1-3
  give the full model and Equations 2, 4 and 5 the simplification; Table 2 gives
  a complete parameter set for each. Fully specified, so fully extracted.
* **Case example 2 (allopurinol / oxypurinol turnover PKPD, NOT packaged).**
  Section 2.2 fits a urate turnover model to 648 patients and concludes that it
  is *not* internally deterministically identifiable, because oxypurinol's
  elimination rate constant and urate's turnover rate constant are too close to
  separate. The resolution the authors adopt in Section 2.2.5 is "to simplify
  the model structure using an immediate effects PD model" -- which is exactly
  the model already in this library as
  [`Wright_2016_allopurinol`](Wright_2016_allopurinol.html), co-authored by this
  paper's first two authors. The turnover variant is therefore both superseded
  and explicitly reported as unreliable (Table 3: `kout` relative standard error
  125%, between-subject variability on `kout` 257 CV% with its own RSE 239%).
  Two further gaps make it unpackageable as it stands: the oxypurinol PK is
  fixed to individual empirical Bayes estimates ("IPP") rather than a reported
  population PK model, so there is no PK layer in the paper; and Table 3's
  `theta diuretic` = 1.7 is given as a "fractional effect of diuretic" without
  saying which parameter it acts on.
* **The stochastic simulation-estimation scenario in Section 2.2.4 (NOT
  packaged).** A hypothetical one-compartment PK plus turnover PD model used to
  demonstrate the `ke` vs `kout` flip. It reports `ke`, three `kout` settings
  and residual-error magnitudes, but not `Emax`, `ka`, `V` or `Rin`, so it
  cannot be assembled without inventing values.

# Population

The source is a tutorial, so the "population" is thin by design and is recorded
honestly rather than embellished. Case example 1 used **80 subjects** given a
single intravenous bolus into the central compartment, with free antibody and
total target (free target plus antibody-target complex) both measured. Neither
the antibody nor its target is named, and no demographics, age, weight, sex or
region are reported. The species is never stated; the framing ("sparse
sampling, common in mAb clinical trials") implies human.

Two sampling designs are used (Section 2.1.3):

| Design | Sampling times (h post-dose) | Free drug obs | Total target obs |
|---|---|---|---|
| Hypothetical rich (structural identifiability template) | 0, 0.01, 0.2, 0.6, 3, 50, 64, 81, 91 | 640 | 720 |
| Reduced clinical (deterministic identifiability, and the simplified model's estimation design) | 0, 0.5, 3, 5, 12, 24 | 400 | 480 |

The datasets are **simulated, not observed**: the reported observation counts are
exactly the sampling grids multiplied by 80 subjects (80 x 9 = 720 and
80 x 6 = 480 total target; the free-drug counts are one time point fewer,
80 x 8 = 640 and 80 x 5 = 400).

The same information is available programmatically via
`readModelDb("Duffull_2025_mab_tmdd_qss")()$population`.

# Source trace

Per-parameter origins are also recorded as in-file comments beside each `ini()`
entry in `inst/modeldb/pharmacokinetics/Duffull_2025_mab_tmdd_qss.R` and
`..._simplified.R`.

## Equations

| Model element | Encoded as | Source location |
|---|---|---|
| `dLC/dt = -(ke + k12) LC + k21 LT - kint LC Rtot / (LC + Km Vc)` | `d/dt(central)`, full model | Duffull 2025 Equation 1, p. 1551 |
| `dLT/dt = k12 LC - k21 LT` | `d/dt(peripheral1)`, both models | Duffull 2025 Equation 2, p. 1551 |
| `dRtot/dt = -(kint - kdeg) LC Rtot / (LC + Km Vc) - kdeg Rtot + ksyn Vc` | `d/dt(total_target)`, full model | Duffull 2025 Equation 3, p. 1551 |
| `dLC/dt = -(ke + k12) LC + k21 LT - kint Rtot` | `d/dt(central)`, simplified model | Duffull 2025 Equation 4, p. 1551 |
| `dRtot/dt = -kint Rtot + ksyn VC` | `d/dt(total_target)`, simplified model | Duffull 2025 Equation 5, p. 1551 |
| `LC(0) = 0, LT(0) = 0, Rtot(0) = ksyn VC / kdeg` | state initial conditions | Duffull 2025, stated below Equation 5, p. 1551 |
| `ke = CL/Vc`, `k12 = Q/Vc`, `k21 = Q/Vp` | micro-constants in `model()` | Derived: the equations are written in rate constants but Table 2 tabulates CL, Vc, Vp, Q |
| Two compartments, single IV bolus into central, binding in central only | model structure | Duffull 2025 Section 2.1.5, p. 1551 |
| Outputs: free drug concentration, total target concentration | `Cc`, `Ctotal_target` | Duffull 2025 Section 2.1.1, p. 1550 |

## Parameters

`Full` is the Table 2 "Nominal value" column (the parameter set at which the
Fisher information matrix was evaluated); `Simplified` is the Table 2
"Simplified model" column (final estimates, RSE in parentheses).

| Parameter | Full | Simplified | Source location |
|---|---|---|---|
| `lcl` (CL, L/h) | 4.13 | 4.15 (6%) | Duffull 2025 Table 2 row CL |
| `lvc` (Vc, L) | 76.7 | 68.4 (9%) | Duffull 2025 Table 2 row Vc |
| `lvp` (Vp, L) | 70 | 75.8 (5%) | Duffull 2025 Table 2 row Vp |
| `lq` (Q, L/h) | 44.7 | 59.9 (11%) | Duffull 2025 Table 2 row Q |
| `lksyn` (ksyn, nmol/L/h) | 1.09 | 1.16 (6%) | Duffull 2025 Table 2 row ksyn |
| `lkdeg` (kdeg, 1/h) | 0.349 | 0.373 (4%) | Duffull 2025 Table 2 row kdeg |
| `lkint` (kint, 1/h) | 11.1 | 11.5 (5%) | Duffull 2025 Table 2 row kint |
| `lkss` (Km in the source; Kss, nmol/L) | 0.13 | absent | Duffull 2025 Table 2 row Km (dash in the simplified column) |
| `etalcl` (IIV CL, CV%) | 25 | 21.9 (15%) | Duffull 2025 Table 2 row IIV-CL |
| `etalvc` (IIV Vc, CV%) | 25 | 31.1 (17%) | Duffull 2025 Table 2 row IIV-Vc |
| `etalksyn` (IIV ksyn, CV%) | 25 | 32.6 (14%) | Duffull 2025 Table 2 row IIV-ksyn |
| `etalkint` (IIV kint, CV%) | 25 | 28.7 (13%) | Duffull 2025 Table 2 row IIV-kint |
| `propSd`, `propSd_Ctotal_target` | `fixed(0)` | `fixed(0)` | Not reported by the source (see Assumptions) |
| Dose = 10000 nmol IV bolus | vignette event table | **Read off Duffull 2025 Figure 3** (free-drug amount at t = 0); not stated in the text |

Two naming translations were applied, both to existing canonicals:

* The source's `Km` is recorded as `lkss`. Its role in Equations 1 and 3 is the
  quasi-steady-state binding constant of the Gibiansky 2008 TMDD reduction (it
  scales free-drug saturation of target-mediated internalisation while the total
  target remains a dynamic state), which is what this library's
  `PK_2cmt_tmdd_qss` archetype calls `lkss`. It is not a metabolic
  Michaelis-Menten constant (canonical `lkm`), which would describe saturable
  elimination with no target state.
* The source's `Rtot` state is the canonical compartment `total_target`, and its
  concentration output is `Ctotal_target`.

```{r models}
mod_full <- readModelDb("Duffull_2025_mab_tmdd_qss")
mod_simp <- readModelDb("Duffull_2025_mab_tmdd_simplified")

DOSE  <- 10000  # nmol; read off Figure 3
TOBS  <- c(0, 0.5, 3, 5, 12, 24)  # reduced clinical design, Section 2.1.3

# Event-table helper. Both models expose two `~` endpoints (Cc and
# Ctotal_target), so observation rows must carry `dvid` rather than a `cmt`
# naming an algebraic observable; rxode2 returns BOTH observable columns at
# every observation row regardless.
make_events <- function(n, times, id_offset = 0L) {
  ids <- id_offset + seq_len(n)
  dplyr::bind_rows(
    tibble(id = ids, time = 0, amt = DOSE, evid = 1L,
           cmt = "central", dvid = NA_integer_),
    tidyr::expand_grid(id = ids, time = times) |>
      mutate(amt = NA_real_, evid = 0L, cmt = NA_character_, dvid = 1L)
  ) |>
    arrange(id, time, desc(evid))
}
```

# Structural verification

Three checks establish that the packaged ODEs are the source's ODEs, before any
attempt to reproduce a figure. They are the gates that a visual profile
comparison would not catch.

## The simplification is algebraically exact in the saturated limit

Duffull 2025 derives Equations 4 and 5 by setting the saturation fraction
`LC/(LC + Km*Vc)` to 1. That substitution should be exact, and in the target
equation the `kdeg` terms should cancel:

`-(kint - kdeg) Rtot - kdeg Rtot + ksyn Vc = -kint Rtot + ksyn Vc`

So driving `Kss` to zero in the **full** model must reproduce the
**simplified** model run with the same parameters, to solver precision. If the
transcription of either equation set were wrong, this would not hold.

```{r exactness}
ev1 <- make_events(1, seq(0, 24, by = 0.05))
full_pars <- c(lcl = log(4.13), lvc = log(76.7), lvp = log(70), lq = log(44.7),
               lksyn = log(1.09), lkdeg = log(0.349), lkint = log(11.1))

sat_limit <- rxSolve(zeroRe(mod_full), ev1,
                     params = c(full_pars, lkss = log(0.13 * 1e-6)),
                     returnType = "data.frame")
simp_same <- rxSolve(zeroRe(mod_simp), ev1, params = full_pars,
                     returnType = "data.frame")

rel_diff <- max(abs(simp_same$central - sat_limit$central) /
                  pmax(sat_limit$central, 1e-8))
cat(sprintf("max relative difference in free-drug amount = %.3g\n", rel_diff))
stopifnot(rel_diff < 1e-6)
```

## The target-mediated term actually reaches the integrator

In the nlmixr2 model-function form, a named intermediate referencing an ODE
state can silently evaluate to zero inside `d/dt()`, deleting a whole
elimination pathway with no error. Both models therefore write the saturation
fraction inline. This exaggeration test proves the term is live: making `Kss`
enormous switches target-mediated elimination off, which must *raise* exposure.

```{r exaggeration}
nominal <- rxSolve(zeroRe(mod_full), ev1, returnType = "data.frame")
inert   <- rxSolve(zeroRe(mod_full), ev1,
                   params = c(lkss = log(0.13 * 1e6)), returnType = "data.frame")

lc24_nominal <- nominal$central[nrow(nominal)]
lc24_inert   <- inert$central[nrow(inert)]
cat(sprintf("LC(24 h) with nominal Kss = %.1f nmol\n", lc24_nominal))
cat(sprintf("LC(24 h) with Kss x 1e6   = %.1f nmol (target term switched off)\n", lc24_inert))
cat(sprintf("ratio = %.3f\n", lc24_inert / lc24_nominal))
stopifnot(lc24_inert / lc24_nominal > 1.05)  # term is live, not inert
```

## Target baseline and saturated plateau

The total target must start at its drug-free steady state `ksyn/kdeg` and, once
the antibody saturates it, fall to the plateau of Equation 5, `ksyn/kint`.

```{r target-levels}
tt_checks <- tibble(
  Quantity = c("Baseline total target Rtot(0)/Vc",
               "Saturated plateau (Equation 5 steady state)"),
  `Closed form` = c("ksyn / kdeg", "ksyn / kint"),
  Expected = c(1.09 / 0.349, 1.09 / 11.1),
  Simulated = c(nominal$Ctotal_target[1], nominal$Ctotal_target[nrow(nominal)])
) |>
  mutate(`% diff` = 100 * (Simulated - Expected) / Expected)
knitr::kable(tt_checks, digits = 4,
             caption = "Total target concentration (nmol/L), full model, typical values.")
stopifnot(max(abs(tt_checks$`% diff`)) < 1)
```

# Virtual cohort

Each model is simulated as one cohort of 100 subjects on the reduced clinical
design, well inside the 200-per-arm cap. Between-subject variability comes from
the model's own `etalcl`, `etalvc`, `etalksyn` and `etalkint`; there are no
covariates to sample, because Duffull 2025 turns covariate effects off for the
identifiability analyses and reports no covariate model.

```{r cohort}
SEED <- 20251009
N <- 100
grid_dense <- sort(unique(c(seq(0, 24, by = 0.25), TOBS)))
events <- make_events(N, grid_dense)
stopifnot(!anyDuplicated(events[events$evid == 0L, c("id", "time")]))
```

# Simulation

The seed is reset immediately before **each** solve, so both models are simulated
on the same draw of `etalcl`, `etalvc`, `etalksyn` and `etalkint`. That makes the
full-versus-simplified comparison paired at the subject level; seeding once and
solving twice would give each model a different cohort and confound the
structural comparison with sampling noise.

```{r simulate}
set.seed(SEED)
sim_full <- rxSolve(mod_full, events, returnType = "data.frame") |>
  mutate(model = "Full TMDD (Eq 1-3)")
set.seed(SEED)
sim_simp <- rxSolve(mod_simp, events, returnType = "data.frame") |>
  mutate(model = "Simplified (Eq 2, 4, 5)")
sim <- bind_rows(sim_full, sim_simp) |>
  mutate(model = factor(model, levels = c("Full TMDD (Eq 1-3)",
                                          "Simplified (Eq 2, 4, 5)")))
```

# Replicate Figure 3

Figure 3 of Duffull 2025 plots the free-drug amount in the central compartment
(red) against `Km * Vc` (blue) over 24 h on a log scale, with 95% intervals, to
justify the simplification. Both quantities carry between-subject variability
here because `Vc` has an eta.

```{r figure3, fig.cap = "Replicates Figure 3 of Duffull 2025: free-drug amount in the central compartment versus Km * Vc over the first 24 h (full model). Lines are medians, ribbons 95% intervals."}
fig3 <- sim_full |>
  mutate(km_vc = kss * vc) |>
  group_by(time) |>
  summarise(
    drug_med = median(central), drug_lo = quantile(central, 0.025),
    drug_hi  = quantile(central, 0.975),
    km_med   = median(km_vc),   km_lo = quantile(km_vc, 0.025),
    km_hi    = quantile(km_vc, 0.975),
    .groups = "drop"
  )

ggplot(fig3, aes(x = time)) +
  geom_ribbon(aes(ymin = drug_lo, ymax = drug_hi), fill = "red", alpha = 0.15) +
  geom_line(aes(y = drug_med, colour = "Free drug in central compartment"), linewidth = 0.9) +
  geom_ribbon(aes(ymin = km_lo, ymax = km_hi), fill = "blue", alpha = 0.15) +
  geom_line(aes(y = km_med, colour = "Km * Vc"), linewidth = 0.9) +
  scale_y_log10(limits = c(1, 20000)) +
  scale_colour_manual(values = c("Free drug in central compartment" = "red",
                                 "Km * Vc" = "blue")) +
  labs(x = "Time (h)", y = "Amount (nmol)", colour = NULL) +
  theme_bw() + theme(legend.position = "top")
```

The source's justification is that "during the first 24 h, the drug amount in
the central compartment exceeded the `Km x VC` value by almost 100 fold". Checked
against the median profile that Figure 3 draws, the margin holds with room to
spare -- but the cohort tail is worth reporting, because the approximation is a
per-subject claim and the paper's figure only shows the aggregate:

```{r saturation-margin}
ratios <- sim_full |>
  filter(time > 0) |>
  mutate(ratio = central / (kss * vc))

median_margin <- ratios |>
  group_by(time) |>
  summarise(m = median(central) / median(kss * vc), .groups = "drop") |>
  pull(m) |>
  min()

per_subject_min <- ratios |>
  group_by(id) |>
  summarise(mn = min(ratio), .groups = "drop")

cat(sprintf("median-profile minimum free-drug / (Km * Vc) over 0-24 h = %.1f\n",
            median_margin))
cat(sprintf("per-subject minima: median %.1f, 2.5th percentile %.1f, lowest %.1f\n",
            median(per_subject_min$mn), quantile(per_subject_min$mn, 0.025),
            min(per_subject_min$mn)))
cat(sprintf("subjects dipping below 100-fold at some point: %d of %d\n",
            sum(per_subject_min$mn < 100), nrow(per_subject_min)))
stopifnot(median_margin > 100)  # the source's claim, on the profile it plots
```

So the source's stated margin is a property of the typical profile. With the
published 25% CV on `CL` and `Vc`, a minority of subjects (6 of 100 at this seed)
spend part of the window under a 100-fold margin, where the saturated
approximation is correspondingly weaker. That is consistent with -- and slightly
sharpens -- the tutorial's argument: the simplification is justified on the
aggregate design, and the residual error it introduces is concentrated in the
tail.

# Quantitative gates against closed-form identities

The source reports no NCA table, so there is nothing to transcribe for a
side-by-side comparison. Closed-form identities are used instead, and they are
the stronger check: each one gates the ODE encoding, the dose encoding, the
observation scaling and the NCA settings simultaneously.

## Cmax equals dose / Vc

An IV bolus into the central compartment must peak at `dose / Vc` exactly.

```{r cmax-gate}
cmax_gate <- tibble(
  model = c("Full TMDD (Eq 1-3)", "Simplified (Eq 2, 4, 5)"),
  Vc = c(76.7, 68.4),
  Expected = DOSE / Vc,
  Simulated = c(
    rxSolve(zeroRe(mod_full), ev1, returnType = "data.frame")$Cc[1],
    rxSolve(zeroRe(mod_simp), ev1, returnType = "data.frame")$Cc[1]
  )
) |>
  mutate(`% diff` = 100 * (Simulated - Expected) / Expected)
knitr::kable(cmax_gate, digits = 4,
             caption = "Cmax of free antibody (nmol/L), typical values.")
stopifnot(max(abs(cmax_gate$`% diff`)) < 1e-6)
```

## PKNCA instrument gate: AUC0-inf equals dose / CL for the linear reduction

Setting `ksyn` to zero removes the target entirely (`Rtot(0) = ksyn*Vc/kdeg` and
the synthesis term both vanish), so each model collapses to a plain linear
two-compartment IV bolus, for which `AUC0-inf = dose / CL` and the terminal
half-life is `ln(2)` over the smaller eigenvalue of the disposition matrix.
Reproducing both through PKNCA confirms the NCA machinery is set up correctly
before it is used on the target-mediated model.

```{r linear-gate}
ev_long <- make_events(1, sort(unique(c(seq(0, 48, by = 0.25),
                                       seq(48, 336, by = 2)))))

lin_profile <- function(mod) {
  rxSolve(zeroRe(mod), ev_long, params = c(lksyn = log(1e-12)),
          returnType = "data.frame")
}

nca_of <- function(df, intervals) {
  conc <- df |> filter(!is.na(Cc)) |> transmute(id = 1L, time, Cc)
  dose <- tibble(id = 1L, time = 0, amt = DOSE)
  PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(conc, Cc ~ time | id, concu = "nmol/L", timeu = "h"),
    PKNCA::PKNCAdose(dose, amt ~ time | id, doseu = "nmol"),
    intervals = intervals
  ))
}

lin_intervals <- data.frame(start = 0, end = Inf,
                            cmax = TRUE, aucinf.obs = TRUE, half.life = TRUE)

# Analytic terminal half-life of the linear reduction.
analytic_thalf <- function(cl, vc, vp, q) {
  ke <- cl / vc; k12 <- q / vc; k21 <- q / vp
  s <- ke + k12 + k21
  lam2 <- (s - sqrt(s^2 - 4 * ke * k21)) / 2
  log(2) / lam2
}

lin_gate <- bind_rows(
  lapply(
    list(list(nm = "Full TMDD (Eq 1-3)",      mod = mod_full,
              cl = 4.13, vc = 76.7, vp = 70,   q = 44.7),
         list(nm = "Simplified (Eq 2, 4, 5)", mod = mod_simp,
              cl = 4.15, vc = 68.4, vp = 75.8, q = 59.9)),
    function(z) {
      res <- as.data.frame(nca_of(lin_profile(z$mod), lin_intervals)$result)
      getp <- function(p) res$PPORRES[res$PPTESTCD == p][1]
      tibble(
        model = z$nm,
        `AUC0-inf expected (dose/CL)` = DOSE / z$cl,
        `AUC0-inf PKNCA` = getp("aucinf.obs"),
        `t1/2 expected (eigenvalue)` = analytic_thalf(z$cl, z$vc, z$vp, z$q),
        `t1/2 PKNCA` = getp("half.life")
      )
    }
  )
) |>
  mutate(`AUC % diff` = 100 * (`AUC0-inf PKNCA` - `AUC0-inf expected (dose/CL)`) /
           `AUC0-inf expected (dose/CL)`,
         `t1/2 % diff` = 100 * (`t1/2 PKNCA` - `t1/2 expected (eigenvalue)`) /
           `t1/2 expected (eigenvalue)`)

knitr::kable(lin_gate, digits = 3,
             caption = "Linear reduction (ksyn -> 0): PKNCA versus closed form. AUC in nmol*h/L, half-life in h.")
stopifnot(max(abs(lin_gate$`AUC % diff`))  < 0.5)
stopifnot(max(abs(lin_gate$`t1/2 % diff`)) < 2)
```

# PKNCA: AUC0-24 of the target-mediated models

`AUC24h` is the source's own response index -- it is what the Sobol global
sensitivity analysis of Figure 2 attributes variance to -- so it is the natural
exposure metric to summarise. The interval ends at 24 h, which is both the
reduced design's last sampling time and the window over which the
target-saturated approximation was justified.

```{r pknca}
nca_cohort <- function(df, conc_col) {
  conc <- df |>
    filter(!is.na(.data[[conc_col]])) |>
    transmute(id, model, conc = .data[[conc_col]], time)
  dose <- events |>
    filter(evid == 1L) |>
    transmute(id, time, amt) |>
    tidyr::expand_grid(model = unique(df$model)) |>
    select(id, model, time, amt)
  PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(conc, conc ~ time | model + id,
                     concu = "nmol/L", timeu = "h"),
    PKNCA::PKNCAdose(dose, amt ~ time | model + id, doseu = "nmol"),
    intervals = data.frame(start = 0, end = 24,
                           cmax = TRUE, tmax = TRUE, auclast = TRUE)
  ))
}

nca_drug   <- nca_cohort(sim, "Cc")
nca_target <- nca_cohort(sim, "Ctotal_target")

summarise_nca <- function(res, label) {
  as.data.frame(res$result) |>
    filter(PPTESTCD %in% c("cmax", "tmax", "auclast")) |>
    group_by(model, PPTESTCD) |>
    summarise(median = median(PPORRES, na.rm = TRUE),
              p2.5 = quantile(PPORRES, 0.025, na.rm = TRUE),
              p97.5 = quantile(PPORRES, 0.975, na.rm = TRUE), .groups = "drop") |>
    mutate(Analyte = label)
}

nca_tbl <- bind_rows(summarise_nca(nca_drug, "Free antibody (Cc)"),
                     summarise_nca(nca_target, "Total target (Ctotal_target)")) |>
  mutate(Parameter = recode(PPTESTCD,
                            cmax = "Cmax (nmol/L)", tmax = "Tmax (h)",
                            auclast = "AUC0-24 (nmol*h/L)")) |>
  select(Analyte, Parameter, model, median, p2.5, p97.5) |>
  rename("Model" = model, "Median" = median,
         "2.5th pct" = p2.5, "97.5th pct" = p97.5) |>
  arrange(Analyte, Parameter, Model)

knitr::kable(nca_tbl, digits = 3,
             caption = "PKNCA summary over 0-24 h, 100 simulated subjects per model.")
```

## The simplification reproduces the full model over 24 h

This is the tutorial's substantive claim: the simplified model is a usable stand-in
over the observed window. Comparing typical-value profiles with each model's own
published parameters (so the comparison includes the re-estimation, not just the
structural change):

```{r full-vs-simplified}
tv_full <- rxSolve(zeroRe(mod_full), ev1, returnType = "data.frame")
tv_simp <- rxSolve(zeroRe(mod_simp), ev1, returnType = "data.frame")

cmp <- tibble(
  `Time (h)` = TOBS,
  `Full Cc (nmol/L)` = approx(tv_full$time, tv_full$Cc, TOBS)$y,
  `Simplified Cc (nmol/L)` = approx(tv_simp$time, tv_simp$Cc, TOBS)$y
) |>
  mutate(`% diff` = 100 * (`Simplified Cc (nmol/L)` - `Full Cc (nmol/L)`) /
           `Full Cc (nmol/L)`)
knitr::kable(cmp, digits = 3,
             caption = "Typical-value free-antibody concentration, each model with its own Table 2 parameters.")

auc24 <- function(df) {
  d <- df[df$time <= 24, ]
  sum(diff(d$time) * (head(d$Cc, -1) + tail(d$Cc, -1)) / 2)
}
auc_pct <- 100 * (auc24(tv_simp) - auc24(tv_full)) / auc24(tv_full)

cat(sprintf("post-dose sampling times (t > 0): max |%% diff| = %.2f%%\n",
            max(abs(cmp$`% diff`[cmp$`Time (h)` > 0]))))
cat(sprintf("AUC0-24 of the typical profile:   %% diff = %.2f%%\n", auc_pct))
stopifnot(max(abs(cmp$`% diff`[cmp$`Time (h)` > 0])) < 5)
stopifnot(abs(auc_pct) < 5)
```

The one large entry is at `t = 0`, and it is **not** a structural discrepancy --
the simplification was already shown above to be algebraically exact. An IV bolus
starts at `dose / Vc`, so the `t = 0` difference must be exactly the ratio of the
two re-estimated central volumes (76.7 L versus 68.4 L). Checking that closes
the loop:

```{r t0-is-vc-ratio}
t0_observed <- cmp$`% diff`[cmp$`Time (h)` == 0]
t0_expected <- 100 * (76.7 / 68.4 - 1)
cat(sprintf("t = 0 difference observed = %.4f%%, expected from Vc ratio = %.4f%%\n",
            t0_observed, t0_expected))
stopifnot(abs(t0_observed - t0_expected) < 1e-6)
```

So over the observed window the two models agree to within 3% at every post-dose
sampling time and to within 2% on `AUC0-24`, with the remaining difference
attributable to re-estimation of the disposition parameters rather than to the
structural change. That is the tutorial's claim, reproduced.

```{r profiles, fig.cap = "Free antibody (left) and total target (right) over 24 h: full versus simplified model. Lines are medians of 100 subjects, ribbons 95% intervals."}
sim |>
  select(id, time, model, Cc, Ctotal_target) |>
  pivot_longer(c(Cc, Ctotal_target), names_to = "output", values_to = "conc") |>
  mutate(output = recode(output, Cc = "Free antibody",
                         Ctotal_target = "Total target")) |>
  group_by(model, output, time) |>
  summarise(med = median(conc), lo = quantile(conc, 0.025),
            hi = quantile(conc, 0.975), .groups = "drop") |>
  ggplot(aes(time, med, colour = model, fill = model)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~output, scales = "free_y") +
  labs(x = "Time (h)", y = "Concentration (nmol/L)", colour = NULL, fill = NULL) +
  theme_bw() + theme(legend.position = "top")
```

# Limitation of the target-saturated approximation

Setting the saturation fraction to 1 makes target-mediated removal of antibody
`kint * Rtot`, which is **independent of how much antibody is present**. Once
`Rtot` reaches its plateau this is a constant-rate sink, so the free-antibody
state is driven negative when antibody washes out. The full model does not have
this problem, because its saturation fraction goes to zero along with the drug.

This is why the simplified model must not be simulated beyond the window the
source validated.

```{r negative-states}
ev_96 <- make_events(1, seq(0, 96, by = 0.25))
tail_simp <- rxSolve(zeroRe(mod_simp), ev_96, returnType = "data.frame")
tail_full <- rxSolve(zeroRe(mod_full), ev_96, returnType = "data.frame")

first_neg <- tail_simp$time[which(tail_simp$central < 0)][1]
cat(sprintf("simplified model: free-drug amount first goes negative at %.1f h\n", first_neg))
cat(sprintf("simplified model: free-drug amount at 96 h = %.1f nmol\n",
            tail_simp$central[nrow(tail_simp)]))
cat(sprintf("full model:       free-drug amount at 96 h = %.3f nmol (stays non-negative)\n",
            tail_full$central[nrow(tail_full)]))
stopifnot(first_neg > 24)                             # 24 h window is safe
stopifnot(min(tail_full$central) >= -1e-6)            # full model does not
```

Within the 24 h window the same mechanism can still bite for subjects in the
upper tail of `ksyn` and `kint`, where the constant sink is largest. Whether any
subject is affected depends on the eta draw, so it is demonstrated across several
seeds rather than left to chance. It is **not** a solver artifact: the minima are
reproduced to five significant figures at `atol = rtol = 1e-12`.

```{r negative-cohort}
neg_scan <- bind_rows(lapply(c(1, 3, 42, 777, SEED), function(sd) {
  bind_rows(lapply(list(list(nm = "Full TMDD (Eq 1-3)", m = mod_full),
                        list(nm = "Simplified (Eq 2, 4, 5)", m = mod_simp)),
    function(z) {
      set.seed(sd); d  <- rxSolve(z$m, events, returnType = "data.frame")
      set.seed(sd); dt <- rxSolve(z$m, events, returnType = "data.frame",
                                  atol = 1e-12, rtol = 1e-12)
      tibble(Seed = sd, Model = z$nm,
             `Subjects with negative free drug` = n_distinct(d$id[d$central < 0]),
             `Minimum free drug (nmol)` = min(d$central),
             `Minimum at atol=rtol=1e-12` = min(dt$central))
    }))
}))
knitr::kable(neg_scan, digits = 3,
             caption = "Negative free-drug states within the 24 h window, 100 subjects per model per seed.")

# The full model must never go negative; the simplified one may, for some draws.
stopifnot(all(neg_scan$`Subjects with negative free drug`[
  neg_scan$Model == "Full TMDD (Eq 1-3)"] == 0))
# Tolerance-independence: default and 1e-12 minima agree.
stopifnot(max(abs(neg_scan$`Minimum free drug (nmol)` -
                    neg_scan$`Minimum at atol=rtol=1e-12`)) < 1e-2)
```

The cohort used for the figures and the NCA above (seed `r SEED`, the same draw
for both models) has no negative states, so the PKNCA results are not affected.
A user re-running with a different seed may see a small number of affected
subjects in the simplified model and should filter or shorten the window
accordingly.

# Assumptions and deviations

* **Residual error is not reported and is fixed at zero.** Table 2 has no sigma
  row for either output, and no residual-error model is described for case
  example 1. The only sigma values in the paper -- `1E-3` in Section 1.2.1 and
  `0.0225` (15% CV) in Section 1.2.2 -- are illustrative advice for setting up a
  `$DESIGN` run, not estimates for this model, and are deliberately not used.
  Both models are therefore typical-value / IIV-only simulation models
  (`propSd` and `propSd_Ctotal_target` are `fixed(0)`). No variance was
  invented.
* **The `%CV` to `omega^2` convention.** Table 2 reports IIV as `IIV-<param>
  (CV%)`. It is encoded as `omega^2 = (CV%/100)^2`, i.e. the reported `%CV`
  taken as the log-scale standard deviation directly, which is the usual NONMEM
  reporting convention for `$OMEGA` inputs and the convention already adopted
  for this author group in `Wright_2016_allopurinol`. The alternative lognormal
  reading `omega^2 = log(1 + CV^2)` would give 0.06062 rather than 0.0625 for
  the 25% nominal case. The source reports no `omega^2` with which to
  discriminate the two; the difference in variance is about 3%.
* **The dose is read off Figure 3, not the text.** The paper states only "a
  single intravenous bolus dose administered into the central compartment". The
  10000 nmol used here is the free-drug amount at `t = 0` on Figure 3. It is
  corroborated independently: the resulting 24 h free-drug amount, 1756 nmol,
  matches Figure 3's red line at 24 h, and `Km * Vc` = 9.97 nmol matches the
  blue line.
* **Species is inferred.** The source never states it. `population$species`
  records "human (implied, not stated)".
* **The upstream `$DESIGN` tutorial is not on disk.** The structure was
  "adapted from a previously published TMDD example in the tutorial for
  `$DESIGN` in NONMEM" (reference [24]). Nothing is inherited from it: Equations
  1-5 and every parameter value used here come from Duffull 2025 itself, so no
  value depends on the unavailable source.
* **The full model's parameters are nominal, not fitted.** Table 2's first
  column is the parameter set at which the Fisher information matrix was
  evaluated. Only the simplified model's column is a converged estimation
  result. Users should treat `Duffull_2025_mab_tmdd_qss` as a structurally
  faithful model with a plausible parameter set, not as a fit to patient data --
  and note that the source's whole point is that `Kss` is not estimable under a
  realistic design.
* **Micro-constants are derived.** Equations 1, 2 and 4 are written in `ke`,
  `k12` and `k21`, but Table 2 tabulates `CL`, `Vc`, `Vp` and `Q`. The
  conversions `ke = CL/Vc`, `k12 = Q/Vc`, `k21 = Q/Vp` are standard and are the
  only reading consistent with both.
* **No covariates.** Section 1.2.1 directs that "covariate effects will need to
  be turned off" for the identifiability analyses, and no covariate model is
  reported, so `covariateData` is empty for both models.
* **Case example 2 is not packaged**; see the reasons in "What this paper is,
  and what is not packaged" above. The library's
  [`Wright_2016_allopurinol`](Wright_2016_allopurinol.html) is the
  immediate-effects model that Section 2.2.5 identifies as the resolution.
* **The source's saturation margin is a typical-profile claim.** Section 2.1.5
  says the drug amount exceeded `Km x VC` "by almost 100 fold". On the median
  profile that Figure 3 plots, the simulated minimum margin over 0-24 h is about
  175-fold (1000-fold at `t = 0`), so the claim holds with room to spare. Once
  the published 25% CV on `CL` and `Vc` is included, however, about 6% of
  subjects spend part of the window below a 100-fold margin (lowest 41-fold at
  the seed used here). This does not contradict the paper -- Figure 3 shows
  medians -- but it does locate where the approximation is weakest, and it is
  reported in the vignette rather than asserted away.

# Errata

No erratum, corrigendum or author correction was located for
doi:10.1002/psp4.70049 at the time of extraction. Supporting Information S1-S3
(the worked `$DESIGN` and empirical-prior examples) are referenced by the paper
but were not present in the open-access package on disk; they illustrate the
NONMEM workflow rather than supply parameter values, and nothing in the two
packaged models depends on them.
