---
title: "Morphine (Duflot 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Morphine (Duflot 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Duflot_2022_morphine"))$reference`
* Description: `r rxode2::rxode(readModelDb("Duflot_2022_morphine"))$description`
* Article: <https://doi.org/10.1002/psp4.12735> (open access)
* Supplement: Supplementary File S1 (Tables S1-S2, model-building BICc tables,
  the Mlxtran code of the final model in Supplementary Material S4, and the
  authors' R script in Supplementary Material S6) and Supplementary File S2
  (the analysis dataset), both on the publisher's page for the article.

Duflot and colleagues gave healthy volunteers a single morphine dose either as
an intravenous bolus or by a 5-minute nebulization through a standard
emergency-department mask, and fitted a joint parent-metabolite model for
morphine, morphine-3-glucuronide (M3G) and morphine-6-glucuronide (M6G) in
Monolix 2020R1. The model is:

* three-compartment morphine disposition parameterised with micro-constants
  (`vc`, `k12`, `k21`, `k13`, `k31`);
* a Savic transit-compartment chain for the nebulized dose (Monolix
  `depot(type = 2, p = F, ka, Mtt, Ktr)`), with bioavailability F = 3.5%;
* morphine elimination entirely through the two glucuronidation outflows
  `ktr1 * Ac` and `ktr2 * Ac`, each feeding a Savic gamma kernel into one
  transit state and then a glucuronide compartment that shares the morphine
  central volume;
* first-order elimination of each glucuronide.

Amounts are in nmol and concentrations in nmol/L (nM), exactly as in the
authors' dataset; time is in minutes. The vignette converts to the paper's
reporting units with the molecular weights given in the Methods (morphine
285.34 g/mol, glucuronides 461.46 g/mol).

```{r units}
mw_morphine <- 285.34
mw_glucuronide <- 461.46
nmol_per_mg <- 1e6 / mw_morphine # 3504.59 nmol per mg, as in Supplementary Material S4
to_ugL_morphine <- mw_morphine / 1000 # nM -> ug/L
to_ugL_glucuronide <- mw_glucuronide / 1000 # nM -> ug/L
```

# Population

Twenty-seven healthy adults (18-60 years, BMI 19-29 kg/m^2) took part in a
parallel-group randomized phase I trial (NCT01975753): 14 received
intravenous morphine (1-5 mg) and 13 received nebulized morphine (3-8 mg),
with the dose chosen by Dixon's up-and-down method. Median age was 25 (i.v.)
and 27 (NEB) years, median weight 71 and 68 kg, and seven participants in each
arm were men (Duflot 2022 Table 1). Blood was sampled up to 305 minutes
after dosing. The same information is stored in the model metadata:

```{r population}
str(rxode2::rxode(readModelDb("Duflot_2022_morphine"))$population)
```

# Source trace

Every structural equation comes from the Mlxtran listing in Supplementary
Material S4 and every value from Duflot 2022 Table 3. In Table 3 an RSE of
'(-)' marks a value the authors fixed; those are wrapped in `fixed()`.

| Model element | Value | Source |
|---|---|---|
| `lfdepot` (F, nebulized) | 0.035, fixed | Table 3 row F |
| `lka` | 0.046 1/min, fixed | Table 3 row ka |
| `lktr` (absorption Ktr) | 1.23 1/min, fixed | Table 3 row ktr |
| `lmtt` (absorption MTT) | 2.35 min | Table 3 row MTT |
| `lvc` | 1.75 L | Table 3 row Vc |
| `lk12`, `lk21` | 0.188, 0.143 (fixed) 1/min | Table 3 rows k12, k21 |
| `lk13`, `lk31` | 0.306 (fixed), 0.010 (fixed) 1/min | Table 3 rows k13, k31 |
| `lktr_m3g`, `lmtt_m3g` | 0.642 1/min, 8.16 min (fixed) | Table 3 rows ktr1, MTT 1 |
| `lka_m3g` (kam3) | 0.172 1/min | Table 3 row kam3 |
| `lkel_m3g` (km3) | 0.0038 1/min | Table 3 row km3 |
| `lktr_m6g`, `lmtt_m6g` | 0.040 1/min (fixed), 57.5 min | Table 3 rows ktr2, MTT 2 |
| `lka_m6g` (kam6) | 0.186 1/min, fixed | Table 3 row kam6 |
| `lkel_m6g` (km6) | 0.0081 1/min | Table 3 row km6 |
| `e_neb_k13` | 1.39, fixed | Table 3 row k13 (printed 'beta_km3_Route'; see below) |
| `e_neb_kel_m3g` | -1.33, fixed | Table 3 row km3, beta_km3_Route |
| `e_male_kel_m6g` | -0.483, fixed | Table 3 row km6, beta_km6_Sex |
| IIV on MTT, Vc, k12, ktr1, MTT 2, kam3 | omega 0.21, 0.49, 0.61, 0.114, 0.090, 0.942 | Table 3 BSV column (variance = omega^2) |
| IIV block k13 / km3 | omega 0.403 (fixed), 0.18; corr -1 | Table 3 BSV column, row corr_km3_k13 |
| `propSd`, `propSd_m3g`, `propSd_m6g` | 0.29, 0.18, 0.27 | Table 3 residual-error rows |
| `n = MTT * Ktr - 1` per chain | -- | Supplementary Material S4 (N1, N2) and the Monolix `depot()` convention |
| Absorption: Savic kernel into `depot`, then `ka` | -- | Supplementary Material S4 `depot(type=2, ...)` |
| Morphine ODEs (`central`, `peripheral1`, `peripheral2`) | -- | Supplementary Material S4 `ddt_Ac`, `ddt_Ap1`, `ddt_Ap2` |
| Glucuronide formation `Ac * kernel(t)`, transit, elimination | -- | Supplementary Material S4 `ddt_AM3Gtr`, `ddt_AM3G`, `ddt_AM6Gtr`, `ddt_AM6G` |
| Glucuronide volumes = Vc | -- | Table 3 footnote; S4 `M3G = AM3G/V` |
| Covariate form log(theta) = log(theta_pop) + beta | -- | Methods 'Pharmacokinetic modeling' |

```{r model}
mod <- readModelDb("Duflot_2022_morphine")
ui <- rxode2::rxode2(mod)
```

# Nebulized-dose mass balance

The nebulized dose is delivered by the analytic Savic kernel into `depot`
(`f(depot) = 0` removes the ordinary bolus). Every nebulized molecule leaves
`depot` at rate `ka`, so `ka * AUC(depot)` must equal `F * dose` once the
depot is empty. This checks the kernel's normalisation and its
bioavailability placement; both sides use the same parameters, so the
tolerance is tight.

```{r mass-balance}
dose_nmol <- 6.2 * nmol_per_mg
ev_mb <- data.frame(
  id = 1L,
  time = c(0, seq(0.05, 600, by = 0.05)),
  amt = c(dose_nmol, rep(0, 12000)),
  cmt = c("depot", rep(NA_character_, 12000)),
  evid = c(1L, rep(0L, 12000)),
  dvid = c(NA_integer_, rep(1L, 12000)),
  ROUTE_IV = 0L,
  SEXF = 1L
)
sim_mb <- as.data.frame(rxode2::rxSolve(rxode2::zeroRe(ui), ev_mb, returnType = "data.frame"))
auc_depot <- sum(diff(sim_mb$time) * (head(sim_mb$depot, -1) + tail(sim_mb$depot, -1)) / 2)
delivered <- 0.046 * auc_depot
c(delivered_nmol = delivered, expected_nmol = 0.035 * dose_nmol)
stopifnot(
  max(sim_mb$Cc) > 0,
  abs(delivered / (0.035 * dose_nmol) - 1) < 0.01
)
```

# Single-dose simulation against the observed concentrations

Table 2 of the paper lists the observed median dose-normalized concentrations
at each nominal sampling time. A cohort of 200 participants per route, half
of them women, receives the median dose of its arm (3.0 mg i.v., 6.2 mg NEB;
Table 1), and the cohort median of the dose-normalized concentration is
compared with the Table 2 medians.

```{r single-dose-cohort}
rxode2::rxSetSeed(20221)
n_per_arm <- 200
sample_times <- c(5, 7, 15, 35, 65, 125, 185, 245, 305)
grid_single <- sort(unique(c(seq(0, 20, by = 0.25), seq(21, 305, by = 1), sample_times)))

make_single <- function(n, route_iv, dose_mg, id_offset) {
  ids <- id_offset + seq_len(n)
  sexf <- rep(c(1L, 0L), length.out = n)
  doses <- data.frame(
    id = ids, time = 0, amt = dose_mg * nmol_per_mg,
    cmt = if (route_iv == 1L) "central" else "depot",
    evid = 1L, dvid = NA_integer_, ROUTE_IV = route_iv, SEXF = sexf,
    dose_mg = dose_mg
  )
  obs <- tidyr::expand_grid(id = ids, time = grid_single) |>
    dplyr::mutate(
      amt = 0, cmt = NA_character_, evid = 0L, dvid = 1L,
      ROUTE_IV = route_iv, SEXF = sexf[match(id, ids)], dose_mg = dose_mg
    )
  dplyr::bind_rows(doses, obs) |> dplyr::arrange(id, time, dplyr::desc(evid))
}

events_single <- dplyr::bind_rows(
  make_single(n_per_arm, 1L, 3.0, 0L),
  make_single(n_per_arm, 0L, 6.2, n_per_arm)
)
stopifnot(!anyDuplicated(unique(events_single[, c("id", "time", "evid")])))

sim_single <- rxode2::rxSolve(
  ui, events_single,
  keep = c("ROUTE_IV", "dose_mg"), returnType = "data.frame"
) |>
  as.data.frame() |>
  dplyr::mutate(
    treatment = ifelse(ROUTE_IV == 1, "Intravenous", "Nebulized"),
    morphine = Cc * to_ugL_morphine,
    m3g = Cc_m3g * to_ugL_glucuronide,
    m6g = Cc_m6g * to_ugL_glucuronide
  )
stopifnot(
  max(sim_single$morphine[sim_single$ROUTE_IV == 0]) > 0,
  max(sim_single$m3g[sim_single$ROUTE_IV == 1]) > 0
)
```

```{r single-dose-table}
# Duflot 2022 Table 2, dose-normalized medians (ug/L per mg). The M6G rows of
# the nebulized arm are omitted: most values were below the limit of
# quantification and the printed medians are 0.
observed_dn <- tibble::tribble(
  ~treatment, ~time, ~morphine, ~m3g, ~m6g,
  "Intravenous", 5, 12.6, 2.0, NA,
  "Intravenous", 7, 8.2, 4.2, NA,
  "Intravenous", 15, 4.6, 12.9, 0.8,
  "Intravenous", 35, 2.3, 17.4, 2.1,
  "Intravenous", 65, 1.5, 16.6, 2.2,
  "Intravenous", 125, 0.9, 13.4, 1.8,
  "Intravenous", 185, 0.7, 11.0, 1.3,
  "Intravenous", 245, 0.5, 8.6, 1.0,
  "Intravenous", 305, 0.3, 7.0, 0.8,
  "Nebulized", 5, 0.27, NA, NA,
  "Nebulized", 7, 0.29, NA, NA,
  "Nebulized", 15, 0.17, 0.42, NA,
  "Nebulized", 35, 0.12, 0.81, NA,
  "Nebulized", 65, 0.07, 1.12, NA,
  "Nebulized", 125, 0.05, 0.95, NA,
  "Nebulized", 185, 0.03, 0.89, NA,
  "Nebulized", 245, 0.03, 0.70, NA,
  "Nebulized", 305, 0.02, 0.59, NA
) |>
  tidyr::pivot_longer(c(morphine, m3g, m6g), names_to = "analyte", values_to = "observed") |>
  dplyr::filter(!is.na(observed))

simulated_dn <- sim_single |>
  dplyr::filter(time %in% sample_times) |>
  dplyr::mutate(dplyr::across(c(morphine, m3g, m6g), \(x) x / dose_mg)) |>
  tidyr::pivot_longer(c(morphine, m3g, m6g), names_to = "analyte", values_to = "value") |>
  dplyr::group_by(treatment, time, analyte) |>
  dplyr::summarise(simulated = median(value), .groups = "drop")

single_cmp <- observed_dn |>
  dplyr::inner_join(simulated_dn, by = c("treatment", "time", "analyte")) |>
  dplyr::mutate(ratio = simulated / observed)

single_cmp |>
  dplyr::mutate(analyte = factor(analyte, c("morphine", "m3g", "m6g"), c("Morphine", "M3G", "M6G"))) |>
  dplyr::arrange(treatment, analyte, time) |>
  dplyr::rename(
    "Route" = treatment,
    "Analyte" = analyte,
    "Time (min)" = time,
    "Observed median (ug/L per mg)" = observed,
    "Simulated median (ug/L per mg)" = simulated,
    "Simulated / observed" = ratio
  ) |>
  knitr::kable(digits = 3, caption = "Dose-normalized concentrations: cohort median vs. Duflot 2022 Table 2.")

ratio_summary <- single_cmp |>
  dplyr::group_by(treatment, analyte) |>
  dplyr::summarise(
    median_ratio = median(ratio),
    min_ratio = min(ratio),
    max_ratio = max(ratio),
    .groups = "drop"
  )
ratio_summary |>
  dplyr::rename(
    "Route" = treatment,
    "Analyte" = analyte,
    "Median simulated / observed" = median_ratio,
    "Lowest" = min_ratio,
    "Highest" = max_ratio
  ) |>
  knitr::kable(digits = 2, caption = "Simulated/observed ratio across sampling times.")
ratio_range <- function(route, an) {
  r <- ratio_summary[ratio_summary$treatment == route & ratio_summary$analyte == an, ]
  sprintf("%.2f-%.2f", r$min_ratio, r$max_ratio)
}
```

Across the sampling times the simulated median is
`r ratio_range("Intravenous", "morphine")` times the observed morphine median
after i.v. dosing and `r ratio_range("Nebulized", "morphine")` times after
nebulization, and `r ratio_range("Intravenous", "m6g")` times the observed
i.v. M6G median. M3G sits below the observed medians in both routes
(`r ratio_range("Intravenous", "m3g")` i.v., `r ratio_range("Nebulized", "m3g")`
NEB; see the NCA comparison below and *Assumptions and deviations*).

```{r single-dose-assert}
morphine_ratio <- ratio_summary$median_ratio[ratio_summary$analyte == "morphine"]
m6g_ratio <- ratio_summary$median_ratio[ratio_summary$analyte == "m6g"]
stopifnot(
  # A mis-transcribed volume, rate constant or bioavailability moves the whole
  # profile by far more than this.
  all(morphine_ratio > 0.6 & morphine_ratio < 1.5),
  all(m6g_ratio > 0.6 & m6g_ratio < 1.5)
)
```

# PKNCA validation

The dose-normalized AUCs of Table 2 were computed by the authors from the
fitted model: Supplementary Material S6 simulates every participant with
their individual (conditional-mode) parameters and integrates each analyte
to 305 minutes. The comparison below computes the same AUC0-305 per mg with
PKNCA on the simulated cohort above, one PKNCA call per analyte, grouped by
route.

```{r pknca}
dose_df <- events_single |>
  dplyr::filter(evid == 1) |>
  dplyr::transmute(
    id, time, amt = dose_mg,
    treatment = ifelse(ROUTE_IV == 1, "Intravenous", "Nebulized")
  )
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)
intervals <- data.frame(start = 0, end = 305, auclast = TRUE, auclast.dn = TRUE)

run_nca <- function(analyte) {
  conc <- sim_single |>
    dplyr::mutate(Cc = .data[[analyte]]) |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::select(id, time, Cc, treatment)
  conc <- dplyr::bind_rows(
    conc,
    conc |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
  ) |>
    dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
    dplyr::arrange(id, treatment, time)
  conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ time | treatment + id)
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
  as.data.frame(res$result) |> dplyr::mutate(analyte = analyte)
}
nca_all <- dplyr::bind_rows(lapply(c("morphine", "m3g", "m6g"), run_nca))
```

```{r nca-compare}
# Duflot 2022 Table 2, 'AUC (dose-normalized), (ug min/L)'.
published_auc <- tibble::tribble(
  ~analyte, ~treatment, ~auclast.dn,
  "morphine", "Intravenous", 1044,
  "morphine", "Nebulized", 19,
  "m3g", "Intravenous", 3752,
  "m3g", "Nebulized", 245,
  "m6g", "Intravenous", 466,
  "m6g", "Nebulized", 28
)

nca_cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_all |> dplyr::filter(PPTESTCD == "auclast.dn"),
  reference = published_auc,
  by = c("analyte", "treatment"),
  units = c(auclast.dn = "ug*min/L per mg"),
  tolerance_pct = 20
)
knitr::kable(
  nca_cmp,
  caption = "Simulated (PKNCA, cohort median) vs. published dose-normalized AUC0-305. * differs from reference by >20%."
)

auc_ratio <- nca_all |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(id, treatment, analyte, PPORRES) |>
  tidyr::pivot_wider(names_from = analyte, values_from = PPORRES) |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    sim_m3g_morphine = median(m3g / morphine),
    sim_m6g_morphine = median(m6g / morphine)
  ) |>
  dplyr::mutate(
    pub_m3g_morphine = c(4.46, 12.47),
    pub_m6g_morphine = c(0.53, 1.54)
  )
auc_ratio |>
  dplyr::rename(
    "Route" = treatment,
    "M3G/morphine, simulated" = sim_m3g_morphine,
    "M3G/morphine, Table 2" = pub_m3g_morphine,
    "M6G/morphine, simulated" = sim_m6g_morphine,
    "M6G/morphine, Table 2" = pub_m6g_morphine
  ) |>
  knitr::kable(digits = 2, caption = "Metabolite-to-morphine AUC0-305 ratios (median).")
```

The morphine and M6G AUCs agree with Table 2 to within about 20% (the i.v.
morphine row sits at the edge of the 20% flag), and the M6G-to-morphine
ratio is reproduced after i.v. dosing and is within about 20% after
nebulization. The M3G AUC is lower than Table 2 by roughly a third after
i.v. dosing and by just under a half after nebulization. Table 2 summarises the 27 participants' individual fits,
whereas this cohort is simulated from the population parameters; the M3G gap
is discussed under *Assumptions and deviations*. The parameters were not
adjusted.

```{r nca-assert}
auc_sim <- nca_all |>
  dplyr::filter(PPTESTCD == "auclast.dn") |>
  dplyr::group_by(analyte, treatment) |>
  dplyr::summarise(sim = median(PPORRES), .groups = "drop") |>
  dplyr::inner_join(published_auc, by = c("analyte", "treatment")) |>
  dplyr::mutate(pct = 100 * (sim - auclast.dn) / auclast.dn)
stopifnot(
  all(abs(auc_sim$pct[auc_sim$analyte %in% c("morphine", "m6g")]) < 35),
  # M3G: the structural check only -- see the narrative above.
  all(abs(auc_sim$pct[auc_sim$analyte == "m3g"]) < 70)
)
```

# Replication of Figure 4 (titration scenario)

Figure 4 of the paper simulates a titration: three nebulizations of 15 mg,
one every 10 minutes (the 5-minute nebulization is entered as a bolus into
the transit depot, as in the source dataset), or three intravenous boluses of
3 mg, one every 5 minutes, in men and women. The figure shows the median with
the 37.5-62.5% and 25-75% bands; 200 participants per arm are simulated here.

```{r fig4-sim}
rxode2::rxSetSeed(20222)
grid_fig4 <- sort(unique(c(seq(0, 50, by = 0.25), seq(51, 300, by = 1))))

make_titration <- function(n, route_iv, sexf, id_offset) {
  ids <- id_offset + seq_len(n)
  dose_times <- if (route_iv == 1L) c(0, 5, 10) else c(0, 10, 20)
  dose_mg <- if (route_iv == 1L) 3 else 15
  doses <- tidyr::expand_grid(id = ids, time = dose_times) |>
    dplyr::mutate(
      amt = dose_mg * nmol_per_mg,
      cmt = if (route_iv == 1L) "central" else "depot",
      evid = 1L, dvid = NA_integer_
    )
  obs <- tidyr::expand_grid(id = ids, time = grid_fig4) |>
    dplyr::mutate(amt = 0, cmt = NA_character_, evid = 0L, dvid = 1L)
  dplyr::bind_rows(doses, obs) |>
    dplyr::mutate(ROUTE_IV = route_iv, SEXF = sexf) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events_fig4 <- dplyr::bind_rows(
  make_titration(n_per_arm, 1L, 1L, 0L),
  make_titration(n_per_arm, 1L, 0L, 1L * n_per_arm),
  make_titration(n_per_arm, 0L, 1L, 2L * n_per_arm),
  make_titration(n_per_arm, 0L, 0L, 3L * n_per_arm)
)
stopifnot(!anyDuplicated(unique(events_fig4[, c("id", "time", "evid")])))

sim_fig4 <- rxode2::rxSolve(
  ui, events_fig4,
  keep = c("ROUTE_IV", "SEXF"), returnType = "data.frame"
) |>
  as.data.frame() |>
  dplyr::mutate(
    arm = paste(ifelse(ROUTE_IV == 1, "IV", "NEB"), ifelse(SEXF == 1, "Female", "Male"), sep = "-"),
    Morphine = Cc * to_ugL_morphine,
    M3G = Cc_m3g * to_ugL_glucuronide,
    M6G = Cc_m6g * to_ugL_glucuronide
  )

bands_fig4 <- sim_fig4 |>
  tidyr::pivot_longer(c(Morphine, M3G, M6G), names_to = "analyte", values_to = "conc") |>
  dplyr::group_by(arm, analyte, time) |>
  dplyr::summarise(
    q25 = quantile(conc, 0.25), q375 = quantile(conc, 0.375),
    median = median(conc),
    q625 = quantile(conc, 0.625), q75 = quantile(conc, 0.75),
    .groups = "drop"
  )
```

```{r fig4-plot, fig.width = 8, fig.height = 9}
plot_fig4 <- function(an, tmax) {
  bands_fig4 |>
    dplyr::filter(analyte == an, time <= tmax) |>
    ggplot(aes(time)) +
    geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#F4A582", alpha = 0.6) +
    geom_ribbon(aes(ymin = q375, ymax = q625), fill = "#D6604D", alpha = 0.6) +
    geom_line(aes(y = median)) +
    facet_wrap(~arm, scales = "free_y") +
    labs(x = "Time (minutes)", y = paste0(an, " (ug/L)")) +
    theme_bw()
}
plot_fig4("Morphine", 50) + ggtitle("Morphine (Figure 4a)")
plot_fig4("M3G", 300) + ggtitle("M3G (Figure 4b)")
plot_fig4("M6G", 300) + ggtitle("M6G (Figure 4c)")
```

Replicates Figure 4 of Duflot 2022. The morphine panels reproduce the
published figure directly: the i.v. peaks of about 1,750 ug/L and the
nebulized medians of about 13 ug/L at 25 minutes and 6.5 ug/L at 50 minutes.

The glucuronide panels of Figure 4 are uniformly lower than this simulation
by a factor of 285.34/461.46 = 0.618, the ratio of the morphine and
glucuronide molecular weights. That is the pattern expected if the published
glucuronide curves were converted from nmol/L to ug/L with the morphine
molecular weight. The deposited dataset points the same way: its glucuronide
concentrations, converted with 461.46 g/mol, reproduce the observed medians
of Table 2 exactly. So the model output is converted with 461.46 g/mol
throughout this vignette, and the table below compares against Figure 4 on
both bases.

```{r fig4-compare}
# Medians digitised by the maintainers from Figure 4 (median line; peak and
# the last plotted time, 50 min for nebulized morphine, 300 min otherwise).
fig4_digitised <- tibble::tribble(
  ~arm, ~analyte, ~metric, ~published,
  "NEB-Female", "Morphine", "peak", 12.8,
  "NEB-Female", "Morphine", "C50", 6.5,
  "NEB-Male", "Morphine", "peak", 12.6,
  "NEB-Male", "Morphine", "C50", 6.4,
  "IV-Female", "M3G", "peak", 255,
  "IV-Female", "M3G", "C300", 103,
  "IV-Male", "M3G", "peak", 262,
  "IV-Male", "M3G", "C300", 104,
  "NEB-Female", "M3G", "peak", 5.4,
  "NEB-Female", "M3G", "C300", 4.4,
  "NEB-Male", "M3G", "peak", 5.15,
  "NEB-Male", "M3G", "C300", 4.1,
  "IV-Female", "M6G", "peak", 25.7,
  "IV-Female", "M6G", "C300", 4.6,
  "IV-Male", "M6G", "peak", 27.3,
  "IV-Male", "M6G", "C300", 10.1,
  "NEB-Female", "M6G", "peak", 4.5,
  "NEB-Female", "M6G", "C300", 0.95,
  "NEB-Male", "M6G", "peak", 5.0,
  "NEB-Male", "M6G", "C300", 2.0
)

sim_metrics <- bands_fig4 |>
  dplyr::group_by(arm, analyte) |>
  dplyr::summarise(
    peak = max(median[time <= 300]),
    C50 = median[time == 50],
    C300 = median[time == 300],
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(c(peak, C50, C300), names_to = "metric", values_to = "simulated")

fig4_cmp <- fig4_digitised |>
  dplyr::inner_join(sim_metrics, by = c("arm", "analyte", "metric")) |>
  dplyr::mutate(
    # Put the simulated glucuronides on the morphine-molecular-weight basis
    # that the published glucuronide curves appear to use.
    simulated_mw_basis = ifelse(analyte == "Morphine", simulated, simulated * mw_morphine / mw_glucuronide),
    pct_diff = 100 * (simulated_mw_basis - published) / published
  )

fig4_cmp |>
  dplyr::rename(
    "Arm" = arm, "Analyte" = analyte, "Metric" = metric,
    "Figure 4 (ug/L)" = published,
    "Simulated (ug/L)" = simulated,
    "Simulated, morphine-MW basis" = simulated_mw_basis,
    "% diff" = pct_diff
  ) |>
  knitr::kable(digits = 2, caption = "Figure 4 medians vs. simulated medians.")

stopifnot(
  # Structural: a mis-transcribed rate constant, volume or covariate effect
  # moves these by tens of percent.
  abs(median(fig4_cmp$pct_diff)) < 10,
  # Envelope, robust to which checkpoints are noisiest to digitise.
  quantile(abs(fig4_cmp$pct_diff), 0.9) < 25
)
```

On the morphine-molecular-weight basis every checkpoint, including the sex
difference in M6G elimination (about two-fold higher M6G in men at 300
minutes), agrees with Figure 4 to within digitisation error.

# Assumptions and deviations

- **Glucuronide formation is anchored to model time zero.** Supplementary
  Material S4 writes the formation kernel as
  `exp(log(Ac) + log(ktr1) + N1*log(ktr1*t) - ktr1*t - LN1fac)`, i.e. the
  current morphine amount times a Savic gamma density evaluated at the model
  time `t`, not at the time after dose. The model transcribes this exactly.
  In the source data every participant received a single dose at time 0, and
  Figure 4 gives all doses within 20 minutes of the first, so the published
  use cases are reproduced. The M3G kernel is effectively zero after about
  30 minutes and the M6G kernel after about 300 minutes; a dose given later
  than that on the same record forms little or no glucuronide. Start the
  record at the first dose, and do not use the model for regimens that extend
  much beyond the published 20-minute titration window.
- **Morphine elimination.** As in S4, morphine leaves the central compartment
  only through the two glucuronidation outflows (`ktr_m3g * central` and
  `ktr_m6g * central`); the kernel scales how much of that outflow appears as
  glucuronide, so the model does not conserve mass between parent and
  metabolites. This is the published structure.
- **k13 route coefficient label.** Table 3 prints the coefficient in the k13
  row as 'beta_km3_Route (=NEB)'. The covariate-building table (S3, 'Route on
  k13') and the Results text (k13 1.30 vs. 0.35 1/min for NEB vs. i.v.;
  0.306 x exp(1.39) = 1.23) identify it as the route effect on k13.
- **Direction of the M3G route effect.** The Results text describes 'a
  significantly higher M3G elimination (0.0038 vs. 0.0010)' in the NEB
  context, but Table 3 gives km3 = 0.0038 1/min for the i.v. reference with
  beta = -1.33 for NEB (0.0010 1/min). The model follows Table 3; the Figure 4
  replication (flat nebulized M3G profile) confirms this reading.
- **Perfect correlation between k13 and km3.** Table 3 reports
  corr_km3_k13 = -1, which makes the covariance block singular. The
  covariance was scaled by 0.99 (correlation -0.99) so the block can be
  sampled; both variances are kept exactly. The paper fixed omega_k13 (RSE
  '(-)') while estimating omega_km3 and the correlation; nlmixr2 cannot fix
  one element of a block, so the block is left unfixed and the fixed status
  is recorded here and in the model comments.
- **Bioavailability transform.** F was logit-normal in Monolix but has no
  random effect and was fixed, so it is stored as `lfdepot = log(0.035)`.
- **Route and sex coding.** The paper codes route 0 = i.v. / 1 = NEB and sex
  0 = women / 1 = men. The model uses the canonical `ROUTE_IV` (1 = i.v.) and
  `SEXF` (1 = female) columns and applies the paper's coefficients to
  `1 - ROUTE_IV` and `1 - SEXF`, so every published value is carried
  unchanged.
- **Nebulization input.** The 5-minute nebulization is a bolus into the
  transit depot, as in the authors' dataset (no infusion duration); the
  transit chain describes the delay.
- **Undefined kernel before a nebulized dose.** The analytic absorption kernel
  depends on the time since the last `depot` dose, which is undefined for an
  intravenous participant; the model holds the depot input at 0 until the
  first nebulized dose.
- **M3G below the observed data at the population level.** The simulated
  median M3G concentrations are about 0.6 (NEB) to 0.75 (i.v.) of the
  observed medians of Table 2, and the simulated AUCs are about a third
  (i.v.) to a half (NEB) below Table 2's model-derived AUCs, while morphine
  and M6G agree. The
  maintainers also compared typical-value predictions with the authors'
  deposited dataset and saw the same ratio (observed M3G about 1.4-1.6 times
  the prediction), whereas the published Figure 4 matches this model exactly
  once its units are accounted for. The published population parameters
  therefore appear to under-predict M3G at the median. The individual fits
  (Figure 2) and the model-derived AUCs of Table 2 use individual parameters,
  which absorb it. No parameter was adjusted.
- **Figure 4 glucuronide units.** See the Figure 4 section: the published
  glucuronide curves are consistent with a conversion using the morphine
  molecular weight.
- **Covariates screened but not retained.** Age, weight and BMI were tested
  (mean-centred, log-transformed) and not retained; they are recorded in
  `covariatesDataExcluded`.
- **Errata.** No correction notice for this article is linked in Europe PMC
  as of 2026-09-30.
