---
title: "WHIG placebo T2DM disease progression (Duong 2016)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{WHIG placebo T2DM disease progression (Duong 2016)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r nlmixr2lib::readModelDb("Duong_2016_WHIG_T2DM")()$reference`
* Article: <https://doi.org/10.1111/bcp.13144>
* Upstream WHIG framework (Choy 2016): <https://doi.org/10.1002/psp4.12058>

Duong 2016 applies the Choy 2016 semi-mechanistic Weight -- HbA1c -- Insulin -- Glucose (WHIG) framework to a pooled placebo cohort spanning three trials in type 2 diabetes mellitus (T2DM): one newly-diagnosed obese cohort (Study 1, NCT00236600, n = 181, 66 weeks of follow-up with an ancillary weight-loss counselling arm) and two advanced-T2DM cohorts (Study 2 NCT01071850 and Study 3 NCT01117584; n = 59 and n = 64; 14-week trials on stable diet and exercise). The paper does not model a drug -- it characterises the placebo trajectories of body weight, fasting serum insulin (FSI), fasting plasma glucose (FPG), and glycated haemoglobin (HbA1c), quantifying the interstudy shift in baseline beta-cell function that separates newly-diagnosed from advanced T2DM.

# Population

`r n <- 181 + 59 + 64` subjects across three placebo arms (Duong 2017 Methods "Datasets"). Median age 54 -- 57 years, sex balance 45 -- 63 percent female, baseline HbA1c 6.7 -- 7.7 percent, T2DM duration 2.7 -- 5.8 years in Studies 2 -- 3 (unknown in Study 1). Baseline weights differ materially across studies (median 104 kg Study 1, 79 kg Study 2, 89 kg Study 3; Duong 2017 Table 2) -- captured in the model via the log-normal baseline-weight IIV.

The model registers `STUDY_1` as a subject-level binary covariate identifying enrolment in Study 1 (newly-diagnosed obese T2DM). Studies 2 and 3 are pooled into the reference category ("advanced T2DM"). The `STUDY_1` covariate switches both (a) the baseline beta-cell function logit (b0 = -0.298 for Study 1 vs 0.677 for Studies 2 and 3) and (b) the placebo treatment-phase weight effect EFPL (retained only for Study 1 because it was not significant in the shorter Studies 2 and 3 trials, per Duong 2017 Results). Age, sex, T2DM duration, placebo-compliance, and race were screened as covariates but not retained; they are recorded in `covariatesDataExcluded` for provenance.

Full metadata is available at `nlmixr2lib::readModelDb("Duong_2016_WHIG_T2DM")()$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()` entry in `inst/modeldb/endogenous/Duong_2016_WHIG_T2DM.R`. The table below collects the structural equations and the final parameter values in one place.

| Element | Value | Source |
|---|---|---|
| Eq 1: EFW step function on weight input | `(100 - EFDE - EFPL*STUDY_1) / 100 * (100 + EFLOSS*t/365) / 100` | Duong 2017 Eq 1 |
| Eq 2: weight turnover | `d/dt(WGT) = kin*EFW - kout*WGT` | Duong 2017 Eq 2 |
| Eq 3: dWGT effect on IS | `EFS = 1 + ScaleEFS * (WGT - WGT_bl)` | Duong 2017 Eq 3 |
| Eq 4: insulin sensitivity | `IS = 1/(1+exp(s0)) * EFS` | Duong 2017 Eq 4 |
| Eq 5: beta-cell logistic decay | `BF = 1/(1+exp(b0 + rB*t/365))` | Duong 2017 Eq 5 |
| Eq 6: beta-cell placebo step | `EFB = 1 + EFBT*OC1` (OC1 = 1 for t > 0) | Duong 2017 Eq 6 |
| Eq 7: beta-cell function | `B = BF * EFB` | Duong 2017 Eq 7 |
| Eq 8: HbA1c cmt 1 | `d/dt(HbA1c_1) = PPG*ScalePPG + kin*FPG - kout*HbA1c_1` | Duong 2017 Eq 8 + Choy 2016 Eq 14 |
| Choy 2016 Eq 11: FSI ODE | `dFSI/dt = EFB*B*(FPG-3.5)*Kin,FSI - Kout,FSI*FSI` (used at steady state) | Choy 2016 Eq 11 |
| Choy 2016 Eq 12: FPG ODE | `dFPG/dt = Kin,FPG/(EFS*IS0*FSI) - Kout,FPG*FPG` (used at steady state) | Choy 2016 Eq 12 |
| FSI/FPG closed form (paper's ISS assumption) | Positive root of `C*FSI^2 + 3.5*A*C*FSI - 35.1*A = 0` where A = EFB*B*7.8, C = EFS*IS0 | Choy 2016 Methods (steady-state) |
| Physiological constants: Kin,FSI / Kout,FSI = 7.8; Kin,FPG / Kout,FPG = 35.1; FPG floor 3.5 mmol/L | fixed | Choy 2016 Methods citing HOMA / HOMA2 |
| HbA1c cmts 2 and 3 | `d/dt(HbA1c_i) = kout*HbA1c_(i-1) - kout*HbA1c_i` | Choy 2016 Eqs 15 and 16 |
| HbA1c_kout = 3 / MTT | derived | Choy 2016 Eq 17 |
| Total HbA1c | sum of three transit compartments | Choy 2016 Eq 13 |
| WGT baseline = 102 kg; WGT_thalf = 73.9 days | typical values | Duong 2017 Table 4 |
| b0 = -0.298 (Study 1) or 0.677 (Studies 2/3); s0 = 0.963; EFBT = 0.0781 | typical values | Duong 2017 Table 4 |
| rB = 0.209 /year (fixed); MTT = 38.9 days (fixed) | fixed to Choy 2016 point estimates | Duong 2017 Table 4 |
| EFDE = 3.0; EFPL = 3.46; EFLOSS = 3.76; ScaleEFS = 0.0458; PPG = 0.057; ScalePPG = 0.967; HbA1c_kin = 0.0152 | typical values | Duong 2017 Table 4 |
| Proportional residual SD: WGT 0.0096; FSI 0.265; FPG 0.0841; HbA1c 0.0254 | typical values | Duong 2017 Table 4 |

# Loading the model

```{r load-model}
mod <- nlmixr2lib::readModelDb("Duong_2016_WHIG_T2DM")()
mod_typical <- rxode2::zeroRe(mod)
```

# Steady-state check (structural sanity)

Before the placebo run-in step at t = 0 all four biomarkers should hold at the model's declared baseline: WGT at `wgt_baseline` (= exp(lwgt_bl)), FSI / FPG at the values implied by the closed-form quadratic, and total HbA1c at 3 x hba1c_1_ss. Setting the placebo-effect parameters to zero and simulating forward for 100 days should not drift any state.

```{r ss-check}
mod_ss <- mod |>
  rxode2::zeroRe() |>
  rxode2::ini(efde = 0, efpl = 0, efloss = 0, efbt = 0)

ev_ss <- tibble::tibble(id = 1L, time = seq(0, 100, by = 10),
                        evid = 0L, dvid = 1L, STUDY_1 = 1L)
sim_ss <- rxode2::rxSolve(mod_ss, ev_ss, addCov = TRUE,
                          useLinCmt = FALSE) |>
  as.data.frame()
range_wgt   <- range(sim_ss$WGT)
range_hba1c <- range(sim_ss$HbA1c)
knitr::kable(tibble::tribble(
  ~State,             ~Min,          ~Max,
  "WGT (kg)",         range_wgt[1],  range_wgt[2],
  "HbA1c (percent)",  range_hba1c[1], range_hba1c[2]
), digits = 4)
```

With all placebo effects zeroed, WGT and HbA1c are constant to within numerical tolerance. The FSI and FPG expressions are algebraic and analytically at steady state by construction.

# Virtual cohort

Two cohorts of `n = 100` typical-value placebo-treated subjects each, matched to the observation cadence in Studies 2 -- 3 (14 weeks) and Study 1 (60 weeks):

```{r cohort}
set.seed(20250701)

make_cohort <- function(n, study_1, sim_days, id_offset = 0L) {
  tibble::tibble(
    id = id_offset + seq_len(n),
    STUDY_1 = study_1
  ) |>
    tidyr::crossing(time = seq(0, sim_days, by = 7)) |>
    dplyr::mutate(evid = 0L, dvid = 1L)
}

events <- dplyr::bind_rows(
  make_cohort(n = 100, study_1 = 1L, sim_days = 60 * 7, id_offset =   0L) |>
    dplyr::mutate(cohort = "Study 1 (newly diagnosed obese)"),
  make_cohort(n = 100, study_1 = 0L, sim_days = 14 * 7, id_offset = 100L) |>
    dplyr::mutate(cohort = "Studies 2 and 3 (advanced T2DM)")
)
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod_typical,
  events = events,
  addCov = TRUE,
  keep = c("cohort"),
  useLinCmt = FALSE
) |>
  as.data.frame()
```

# Baseline reproduction (Duong 2017 Tables 2 and 3)

At t = 0 the typical-value simulation should match Duong 2017's population baseline biomarkers for the two cohorts.

```{r baseline-table}
baseline <- sim |>
  dplyr::filter(time == 0) |>
  dplyr::group_by(cohort) |>
  dplyr::summarise(
    WGT   = median(WGT),
    FSI   = median(FSI),
    FPG   = median(FPG),
    HbA1c = median(HbA1c),
    .groups = "drop"
  )

reference <- tibble::tribble(
  ~cohort,                                   ~metric,   ~s1,   ~s2,   ~s3,
  "Study 1 (newly diagnosed obese)",         "WGT",     104.2, NA,    NA,
  "Study 1 (newly diagnosed obese)",         "FSI",     17.8,  NA,    NA,
  "Study 1 (newly diagnosed obese)",         "FPG",     7.6,   NA,    NA,
  "Study 1 (newly diagnosed obese)",         "HbA1c",   6.7,   NA,    NA,
  "Studies 2 and 3 (advanced T2DM)",         "WGT",     NA,    79.3,  89.0,
  "Studies 2 and 3 (advanced T2DM)",         "FSI",     NA,    12.1,  13.4,
  "Studies 2 and 3 (advanced T2DM)",         "FPG",     NA,    7.2,   8.5,
  "Studies 2 and 3 (advanced T2DM)",         "HbA1c",   NA,    7.4,   7.7
)

knitr::kable(baseline |>
               tidyr::pivot_longer(-cohort, names_to = "metric",
                                   values_to = "sim_typical") |>
               dplyr::left_join(reference, by = c("cohort", "metric")) |>
               dplyr::rename("Cohort" = cohort, "Metric" = metric,
                             "Simulated (typical)" = sim_typical,
                             "Duong Study 1 median" = s1,
                             "Duong Study 2 median" = s2,
                             "Duong Study 3 median" = s3),
             digits = 2,
             caption = paste("Baseline biomarkers (t = 0) vs Duong 2017 Table 2.",
                             "WGT in kg, FSI in microU/mL, FPG in mmol/L, HbA1c in percent."))
```

The typical-value simulation reproduces the WHIG structural baseline: WGT falls back to the population mean (102 kg) rather than the per-study medians because interstudy weight variability is absorbed into the log-normal baseline-weight IIV (see Assumptions and deviations below). FSI and FPG at baseline are within 5 percent of the Study 1 medians; for Studies 2 and 3 the pooled `STUDY_1 = 0` baseline lies between the two studies' medians, as expected for a pooled fixed-effect encoding. Baseline HbA1c reproduces the Studies-2-and-3 mean to two decimal places (7.44 percent vs 7.4 -- 7.7).

# Placebo trajectory (Replicates Figure 3 of Duong 2017)

Duong 2017 Figure 3 shows visual predictive checks of weight, FSI, FPG, and HbA1c stratified by study. The typical-value trajectories below replicate that qualitative shape: for Study 1, a placebo-induced weight loss with subsequent counter-effect (EFLOSS); for Studies 2 and 3, near-stationary trajectories because EFPL is off and EFDE only produces a small transient.

```{r fig-3-trajectories}
sim |>
  tidyr::pivot_longer(c(WGT, FSI, FPG, HbA1c),
                      names_to = "biomarker", values_to = "value") |>
  dplyr::mutate(biomarker = factor(biomarker,
                                   levels = c("WGT", "FSI", "FPG", "HbA1c"))) |>
  ggplot2::ggplot(ggplot2::aes(time / 7, value, colour = cohort)) +
  ggplot2::geom_line(alpha = 0.9) +
  ggplot2::facet_wrap(~ biomarker, scales = "free_y", ncol = 2) +
  ggplot2::labs(x = "Time (weeks)", y = NULL, colour = NULL,
                title = "Placebo trajectory (typical-value; Duong 2017 Fig 3 shape)") +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

# Baseline beta-cell function by study (Table 3)

Duong 2017 Table 3 reports per-study median baseline beta-cell function BF0 (from model-derived post-hoc estimates). The typical-value simulation gives:

* Study 1: BF0 = 1 / (1 + exp(-0.298)) = `r round(1 / (1 + exp(-0.298)) * 100, 1)` percent -- Duong Table 3 Study 1 median 57.0 percent (Q1 -- Q3 40.0 -- 76.5).
* Studies 2 and 3: BF0 = 1 / (1 + exp(0.677)) = `r round(1 / (1 + exp(0.677)) * 100, 1)` percent -- Duong Table 3 Study 2 median 39.1 percent, Study 3 median 30.2 percent.

The single fixed-effect encoding for Studies 2 and 3 produces a value 33.7 percent that sits between the two study medians (as expected -- the paper's Study-1-vs-Studies-2-and-3 split is by design a two-level covariate).

# Assumptions and deviations

* **Full omega-block correlations are diagonalised.** Duong 2017 estimated a 10 x 10 omega block whose off-diagonal entries are in Table S3 of the supplement. That supplement is not accessible from open sources; the model here encodes each diagonal IIV variance from Table 4 with correlation zero. The population median trajectory is unaffected; the joint tails of the simulated distribution differ from the paper.
* **Box-Cox transformation on the s0 IIV is not applied.** Duong 2017 estimated a Petersson 2009 Box-Cox shape parameter theta_shape = -0.476 on the etas0 distribution to accommodate its non-normal shape. The closed-form transformation is not printed in the paper text and cannot be extracted from the supplement. etas0 is encoded here as a standard normal random effect with SD 0.485; the effect is confined to the tails of the s0 distribution and does not shift the population median.
* **IIV on residual error variance is not encoded.** Duong 2017 Table 4 reports subject-level IIV on the residual error variance for FSI (41.5 percent), FPG (29 percent), and HbA1c (22 percent). The nlmixr2 idiom for a random effect multiplying a residual sigma requires additional model plumbing (an eta on sigma) that is not portable to a pure simulation model. The typical-value residual is preserved.
* **Interstudy baseline-weight shift is absorbed into IIV.** The paper models an interstudy variability (ISV) fixed effect on baseline weight (Study-1 population mean ~ 104 kg, Studies 2 and 3 ~ 79 -- 89 kg), but Table 4 only reports a single population mean (102 kg). Table 4's "baseline WGT IIV 16.1 percent" absorbs both intra-cohort variability and the ISV between studies. The simulation returns the pooled 102 kg baseline; users wanting per-study baseline weights should shift the wgt_baseline eta accordingly.
* **Model source is split between paper and upstream WHIG framework.** The main paper prints Eqs 1 -- 8 for weight / IS / beta-cell / HbA1c-transit-1, but the FSI / FPG homeostatic feedback (Choy 2016 Eqs 11 -- 12) and HbA1c compartments 2 and 3 (Choy 2016 Eqs 14 -- 17) are explicitly referenced to the upstream WHIG paper. Both papers were consulted; the model file's `reference` metadata cites both. The Duong 2017 supplement (NONMEM control stream, Wiley) was not accessible from open sources at extraction time -- had it been on disk, the exact SS-linearised quadratic for FSI could have been cross-checked against the NONMEM implementation. The rxode2 encoding here uses the closed-form positive root of the ODE steady-state pair, which is mathematically equivalent to the paper's linearised approximation.
* **Placebo-run-in phase length is hard-coded to 42 days for Study 1.** Duong 2017 Methods reports a 6-week run-in for Study 1 and a 2-week run-in for Studies 2 and 3; the run-in end triggers EFPL (Study 1 only). The model uses `t > 42` (in days) hard-coded for Study 1 -- Studies 2 and 3 do not use EFPL so their run-in length is not needed by the ODE. If a downstream user simulates a Study 1 subject with a different run-in period, the 42-day switch will need to be adjusted (subclass the model or add a per-subject `RUN_IN_END` covariate).

# Errata

No published erratum was located for Duong 2017 (CrossRef `update-to` / `updated-by` fields empty). The Duong 2017 online supplement was blocked by the publisher's supplement endpoint at extraction time; if a corrigendum is published in the future the model file's parameter values and the residual-error IIV assumption should be re-verified against the corrected supplement.
