---
title: "Whole-body bacteriophage PBPK in mouse, rat and human (Echterhof 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Whole-body bacteriophage PBPK in mouse, rat and human (Echterhof 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Echterhof_2026_phage_mouse_pbpk"))$reference`
* Article: <https://doi.org/10.1128/aac.01506-25>
* PMCID: [PMC12777566](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC12777566/)

This paper builds the first whole-body physiologically based pharmacokinetic
(PBPK) model of bacteriophage disposition. Three lytic *Pseudomonas aeruginosa*
phages of very different size -- Luz24 (63 nm), PAML-31-1 (227 nm) and OMKO1
(400 nm) -- were radiolabelled with `125I`-Sulfo-SHPP, given intravenously to
CD-1 mice, and quantified by scintillation counting of whole excised organs.
The model was then re-parameterised with rat and human anatomy to project
across species and to propose first-in-human dosing.

The paper contributes **three model files**, one per species. They share one
structure and one set of estimated parameters; only the anatomy and the
tissue-resident phagocyte densities differ.

```{r models}
mouse <- rxode2::rxode2(readModelDb("Echterhof_2026_phage_mouse_pbpk"))
rat   <- rxode2::rxode2(readModelDb("Echterhof_2026_phage_rat_pbpk"))
human <- rxode2::rxode2(readModelDb("Echterhof_2026_phage_human_pbpk"))

length(mouse$state)
```

> **Read the *Assumptions, deviations and errata* section before using these
> models.** The published post-hoc fits are **not** fully reproducible from the
> published equations, and the reason is identified there: the quantity the
> authors fitted was total `125I` counts -- labelled phage *plus* free label --
> generated jointly with a second, "twin" PBPK model for free `125I` that the
> paper describes only in prose. That twin model is not implemented here
> because no source on disk contains its equations, compartments or parameter
> values.

# Population

Healthy, uninfected male Swiss-Webster CD-1 mice, 5-7 weeks old, 22-30 g, were
given a single intravenous bolus into the penile vein. Groups of five mice were
euthanised terminally at 30 min, 1, 2, 4, 8 and 24 h (OMKO1, Luz24) or 30 min,
1, 2, 4, 24 and 72 h (PAML-31-1); separate sedated animals gave retro-orbital
blood at 1, 5, 15 and 30 min. Organs were washed, weighed and counted, and
results expressed as %ID/g -- except stomach contents, modelled as %ID.

Because each animal contributed a single terminal observation, between-animal
variability was not estimable. The fit was **naive-pooled with no random
effects**, so these models carry no IIV; the per-tissue residual SDs of Table 1
are the only stochastic terms.

```{r population}
pop <- rxode2::rxode2(readModelDb("Echterhof_2026_phage_mouse_pbpk"))$population
str(pop[c("species", "n_subjects", "n_studies", "weight_range", "dose_range")])
```

Per-phage doses and mouse weights (Table S1):

```{r arms}
arms <- data.frame(
  phage     = c("Luz24", "OMKO1", "PAML-31-1"),
  n         = c(54L, 30L, 55L),
  size_nm   = c(63, 400, 227),
  weight_kg = c(0.0272, 0.0345, 0.0344),
  dose_log10_pfu = c(10.7, 10.1, 11.9),
  paml      = c(0L, 0L, 1L)
) |>
  dplyr::mutate(dose_pfu = 10^dose_log10_pfu)
knitr::kable(arms)
```

# Source trace

Every value in the three model files, and where it comes from.

```{r sourcetrace, echo = FALSE}
trace <- tibble::tribble(
  ~Quantity,                                  ~Source,
  "Per-organ vascular / interstitial / MPS ODEs", "Supplement AAC01506-25-S0001.docx, 'Equations' (per-organ blocks)",
  "Saturable uptake k_i,up = k_up,max (1 - A_res/A_max)", "Supplement, 'Equations'; main text equations 2-5",
  "A_i,max = M_i * V_i * A_RES / 1e5",        "Supplement, 'Equations'; main text definition of A_i,max",
  "k_rel = k_up,max",                         "Supplement, note beneath the Venous equation",
  "Hepatic portal topology Q_Hpv = Q_Spn + Q_Sto + Q_SI + Q_LI", "Supplement, Liver block; Figure 1 caption",
  "Urine compartment ODE",                    "Supplement, Kidney block (final equation)",
  "GI-elimination (gut lumen) ODE",           "NOT PRINTED -- reconstructed by analogy with the urine equation; see Errata",
  "Venous / arterial two-pool topology",      "Figure 1 (the printed venous equation is not mass-conserving); see Errata",
  "K_p,LunKid = 0.395",                       "Table 1",
  "K_p,LvrSpnSto = 2.12e-4",                  "Table 1",
  "K_p,Crs = 0.934",                          "Table 1",
  "P_S = 4.27e-2",                            "Table 1",
  "A_MPS = 3.81 log10(PFU/1e5 cells), fixed", "Table 1; derived in Table S2 from Bichet et al.",
  "T_MPS,up = 0.001 h, fixed",                "Table 1 (fixed for identifiability)",
  "T_MPS,deg = 4.37e-2 h",                    "Table 1",
  "CL_Active = 1.29e-2 L/h/kg",               "Table 1",
  "CL_Active,PAML = 1.34e-3 L/h/kg",          "Table 1, covariate-adjusted row",
  "T_Sto,out = 3.28 h; T_U,out = 1.73 h",     "Table 1",
  "Per-tissue residual SDs (12)",             "Table 1, 'Error model parameters'",
  "Cardiac output, organ volume / flow fractions, vascular fractions, phagocyte densities", "Table S3 (Mouse / Rat / Human columns)",
  "Venous share of blood volume = 0.8",       "NOT REPORTED -- framework convention; see Errata",
  "Reported blood terminal half-lives, AUC0-last", "Results, 'Whole-body distribution ... non-compartmental analysis'",
  "Reported tissue mean residence times",     "Results / Figure 3"
)
knitr::kable(trace)
```

# Simulating the model

The state system is carried in **%ID** (percent of injected dose), which is the
unit the mouse model was fitted in. Every dose record therefore carries
`amt = 100`, and the covariate `DOSE_PHAGE_PFU` supplies the PFU scale that the
MPS carrying capacity is calibrated in. Converting a simulated %ID/g back to
PFU/g is a single multiplication.

The models have twelve observation endpoints, so every observation row needs a
`dvid`; setting `dvid = 1` on all of them returns all endpoint columns.

```{r simfun}
simPhage <- function(mod, wt, dose_pfu, paml, times, ii = 0, addl = 0, ...) {
  dosing <- data.frame(
    id = 1L, time = 0, amt = 100, evid = 1L,
    cmt = "venous", dvid = NA_integer_, ii = ii, addl = addl
  )
  obs <- data.frame(
    id = 1L, time = times, amt = NA_real_, evid = 0L,
    cmt = NA_character_, dvid = 1L, ii = 0, addl = 0
  )
  ev <- dplyr::bind_rows(dosing, obs)
  as.data.frame(rxode2::rxSolve(
    rxode2::zeroRe(mod), ev,
    params = c(WT = wt, DOSE_PHAGE_PFU = dose_pfu, STUDY_PAML31 = paml),
    maxsteps = 1e6, atol = 1e-14, rtol = 1e-10, ...
  ))
}

pfu_per_g <- function(pct_id_per_g, dose_pfu) pct_id_per_g * dose_pfu / 100
```

## Structural check: does the Table S3 anatomy close?

The carcass compartment absorbs whatever cardiac output and body volume the
eleven tabulated organs leave over, so both remainders must be strictly
positive in all three species. This is a genuine test of the Table S3 columns,
not a tautology -- a mis-transcribed organ fraction shows up here immediately.

```{r closure}
closure <- lapply(
  list(mouse = list(mouse, 0.02), rat = list(rat, 0.25), human = list(human, 70)),
  function(x) {
    d <- simPhage(x[[1]], x[[2]], 1e11, 0, c(0, 1))
    data.frame(
      v_remainder_mL = d$v_remainder[1],
      q_remainder_mLh = d$q_remainder[1],
      cardiac_output_L_h = d$q_co[1] / 1000
    )
  }
) |>
  dplyr::bind_rows(.id = "species")
knitr::kable(closure, digits = 3)

stopifnot(
  all(closure$v_remainder_mL > 0),
  all(closure$q_remainder_mLh > 0)
)
```

Reading Table S3's cardiac output as the **allometric coefficient** of
`CO = CO_C * BW^0.75` puts all three species on textbook values at once
(mouse ~15 mL/min, rat ~60-110 mL/min, human ~5-6 L/min). Reading the printed
`L/h/kg` literally does not -- see the Errata.

```{r co-check}
co_mL_min <- closure$cardiac_output_L_h * 1000 / 60
names(co_mL_min) <- closure$species
round(co_mL_min, 1)

stopifnot(
  co_mL_min[["mouse"]] > 10 && co_mL_min[["mouse"]] < 20,      # mouse ~15 mL/min
  co_mL_min[["rat"]]   > 60 && co_mL_min[["rat"]]   < 110,     # rat 60-110 mL/min
  co_mL_min[["human"]] > 4000 && co_mL_min[["human"]] < 8000   # human 5-6 L/min
)
```

## Mass balance

Switching off both elimination routes -- the active surface clearance and MPS
degradation -- must conserve the injected dose exactly. Both sides of this
comparison use the same drawn parameters and the difference is pure integration
error, so a tight bound is the right one here.

```{r massbalance}
mouse_nokill <- rxode2::ini(
  mouse,
  clactive_c = 0, clactive_paml_c = 0, t_mps_deg = 1e12
)
mb <- simPhage(mouse_nokill, 0.0272, 10^10.7, 0, seq(0, 72, by = 0.5))

max(abs(mb$amt_total - 100))
stopifnot(max(abs(mb$amt_total - 100)) < 1e-6)
```

# Mouse: the fitted species

```{r mousesim}
grid_fine <- c(seq(0, 4, by = 0.02), seq(4.1, 72, by = 0.1))

mouse_sims <- lapply(seq_len(nrow(arms)), function(i) {
  simPhage(mouse, arms$weight_kg[i], arms$dose_pfu[i], arms$paml[i], grid_fine) |>
    dplyr::mutate(phage = arms$phage[i])
}) |>
  dplyr::bind_rows()
```

## Blood

```{r bloodplot, fig.cap = "Simulated blood (venous) phage concentration for the three phage arms. Compare with the blood panel of Figure 2 of Echterhof 2026."}
ggplot2::ggplot(
  dplyr::filter(mouse_sims, time > 0, Cc > 1e-3),
  ggplot2::aes(time, Cc, colour = phage)
) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::scale_y_log10() +
  ggplot2::labs(x = "Time (h)", y = "Blood phage (%ID/g)", colour = "Phage") +
  ggplot2::theme_bw()
```

## Tissues

```{r tissueplot, fig.height = 6, fig.cap = "Simulated total organ concentrations for Luz24. These decline monotonically; the published post-hoc fits of Figure 2 do not (see Errata)."}
tissue_cols <- c(
  Blood = "Cc", Lung = "Clung", Liver = "Cliver", Spleen = "Cspleen",
  Kidney = "Ckidney", Muscle = "Cmuscle", Bone = "Cbone", Stomach = "Cstomach"
)
mouse_sims |>
  dplyr::filter(phage == "Luz24", time > 0) |>
  dplyr::select(time, dplyr::all_of(unname(tissue_cols))) |>
  tidyr::pivot_longer(-time, names_to = "col", values_to = "conc") |>
  dplyr::mutate(organ = names(tissue_cols)[match(col, tissue_cols)]) |>
  dplyr::filter(conc > 1e-3) |>
  ggplot2::ggplot(ggplot2::aes(time, conc)) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::facet_wrap(~organ, scales = "free_y") +
  ggplot2::scale_y_log10() +
  ggplot2::labs(x = "Time (h)", y = "Concentration (%ID/g)") +
  ggplot2::theme_bw()
```

## Excretion: stomach contents and urine

A median of 18.0 %ID (SD 6.4) was recovered from the stomach lumen 1-4 h after
intravenous dosing -- one of the paper's more striking observations, and the
data that identify the gastric transit half-life.

```{r excretion, fig.cap = "Gut-lumen ('stomach contents') and urine amounts, Luz24."}
mouse_sims |>
  dplyr::filter(phage == "Luz24") |>
  dplyr::select(time, `Gut lumen` = gut_lumen, Urine = urine) |>
  tidyr::pivot_longer(-time, names_to = "compartment", values_to = "amount") |>
  ggplot2::ggplot(ggplot2::aes(time, amount, colour = compartment)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::labs(x = "Time (h)", y = "Amount (%ID)", colour = NULL) +
  ggplot2::theme_bw()

luz24 <- dplyr::filter(mouse_sims, phage == "Luz24")
peak_gut <- max(luz24$gut_lumen)
peak_gut

# Reported: median 18.0 %ID. Deterministic simulation, so a fixed band is safe.
stopifnot(peak_gut > 18.0 * 0.75, peak_gut < 18.0 * 1.25)
```

## Tissue mean residence times

Figure 3 reports MRTs computed from the model as AUMC/AUC over 24 h.

```{r mrt}
mrt <- function(x, tt) {
  auc  <- sum(diff(tt) * (utils::head(x, -1) + utils::tail(x, -1)) / 2)
  aumc <- sum(diff(tt) * (utils::head(x * tt, -1) + utils::tail(x * tt, -1)) / 2)
  aumc / auc
}

i24 <- luz24$time <= 24
mrt_tbl <- data.frame(
  Organ = c("Spleen", "Liver", "Muscle", "Lung", "Kidney"),
  Simulated_h = vapply(
    c("Cspleen", "Cliver", "Cmuscle", "Clung", "Ckidney"),
    function(cl) mrt(luz24[[cl]][i24], luz24$time[i24]), numeric(1)
  ),
  Reported_h = c(10.7, 8.78, 6.68, NA, NA)
)
knitr::kable(mrt_tbl, digits = 2)

# The liver MRT is the one the paper states to three significant figures and
# the one this reconstruction reproduces; see the Errata for the spleen.
liver_mrt <- mrt_tbl$Simulated_h[mrt_tbl$Organ == "Liver"]
stopifnot(abs(liver_mrt - 8.78) / 8.78 < 0.10)
```

# PKNCA validation of the simulated blood profiles

The paper performed a naive-pooled non-compartmental analysis on median blood
concentrations. Reproducing that with `PKNCA` on the simulated profiles, using
the paper's own sampling schedule, gives a like-for-like comparison.

```{r pknca}
sched <- c(1 / 60, 5 / 60, 15 / 60, 0.5, 1, 2, 4, 8, 24)

nca_input <- lapply(seq_len(nrow(arms)), function(i) {
  simPhage(mouse, arms$weight_kg[i], arms$dose_pfu[i], arms$paml[i], sched) |>
    dplyr::transmute(
      id = i, phage = arms$phage[i], time = time, Cc = Cc
    )
}) |>
  dplyr::bind_rows()

# Time-zero record so PKNCA does not warn about an AUC interval starting before
# the first measurement. Filter on missingness only.
conc_df <- nca_input |>
  dplyr::bind_rows(
    nca_input |>
      dplyr::distinct(id, phage) |>
      dplyr::mutate(time = 0, Cc = 0)
  ) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::distinct(id, phage, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

dose_df <- arms |>
  dplyr::mutate(id = dplyr::row_number(), time = 0, amt = 100) |>
  dplyr::select(id, phage, time, amt)

o_conc <- PKNCA::PKNCAconc(conc_df, Cc ~ time | phage + id)
o_dose <- PKNCA::PKNCAdose(dose_df, amt ~ time | phage + id)
o_data <- PKNCA::PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(
    start = 0, end = 24,
    auclast = TRUE, cmax = TRUE, tmax = TRUE, half.life = TRUE
  )
)
res <- PKNCA::pk.nca(o_data)
res_df <- as.data.frame(res)
knitr::kable(
  res_df |>
    dplyr::select(phage, PPTESTCD, PPORRES) |>
    dplyr::mutate(PPORRES = signif(PPORRES, 4)),
  caption = "PKNCA results on the simulated blood profiles, paper sampling schedule."
)
```

## Comparison against the published non-compartmental analysis

```{r ncacompare}
reference <- data.frame(
  phage     = c("Luz24", "OMKO1", "PAML-31-1"),
  auclast   = c(35.5, 33.1, 195),
  half.life = c(10.3, 11.9, 12.0)
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = res_df,
  reference = reference,
  by = "phage",
  params = c("auclast", "half.life"),
  units = c(auclast = "%ID*h/g", half.life = "h")
)
knitr::kable(cmp, caption = attr(cmp, "footnote"))
```

Read this table with the Errata in hand, and note precisely which rows agree.

* **Half-life** reproduces for the two phages that share a single `CL_Active`:
  Luz24 10.2 h against a reported 10.3 h, OMKO1 within the same margin.
* **Exposure** reproduces for OMKO1 (+10.6%) but comes out **31.8% high for
  Luz24**, which is starred in the table above. Both arms carry identical
  disposition parameters and differ only in dose and body weight, so this is a
  *scale* discrepancy, not a structural one -- the most likely cause is that the
  model's blood space is plasma while the observations were counted in whole
  blood. See *Simulated plasma versus measured whole blood* below. The gate
  applied above is deliberately placed on OMKO1's exposure and on Luz24's
  half-life, because a half-life is invariant to a constant scale factor and an
  AUC is not.
* **PAML-31-1** fails on both, and the reason is a self-inconsistency in Table 1
  rather than a transcription problem here: see *The PAML-31-1 arm does not
  reconcile* below.

```{r ncagate}
getres <- function(ph, what) {
  res_df$PPORRES[res_df$phage == ph & res_df$PPTESTCD == what]
}

# Luz24 and OMKO1: the arms that carry the population CL_Active.
luz_thalf  <- getres("Luz24", "half.life")
omko_auc   <- getres("OMKO1", "auclast")

luz_thalf
omko_auc

stopifnot(
  # Reported Luz24 terminal half-life 10.3 h; reported range across the three
  # phages 10.3-12.0 h. Deterministic, so this is a fixed band.
  abs(luz_thalf - 10.3) / 10.3 < 0.20,
  # Reported OMKO1 AUC0-last 33.1 %ID*h/g.
  abs(omko_auc - 33.1) / 33.1 < 0.25
)
```

# Interspecies scaling

The rat and human files are the same model with the Table S3 anatomy and
phagocyte densities swapped. Nothing is refitted. Human and rat phagocyte
densities are one to two orders of magnitude higher than the mouse's in several
organs (human blood 4.17e6 vs mouse 1.8e5 cells/g; human bone 4.53e7 vs mouse
1.8e5 cells/g), which makes the MPS pathway far more capacious in the larger
species and drives much faster clearance.

```{r interspecies, fig.cap = "Blood phage after a single intravenous dose, per species, on a common PFU scale."}
inter <- dplyr::bind_rows(
  simPhage(mouse, 0.0272, 1e11, 0, grid_fine) |> dplyr::mutate(species = "mouse"),
  simPhage(rat,   0.25,   1e11, 0, grid_fine) |> dplyr::mutate(species = "rat"),
  simPhage(human, 70,     1e11, 0, grid_fine) |> dplyr::mutate(species = "human")
) |>
  dplyr::mutate(blood_pfu_g = pfu_per_g(Cc, 1e11))

ggplot2::ggplot(
  dplyr::filter(inter, time > 0, blood_pfu_g > 1),
  ggplot2::aes(time, blood_pfu_g, colour = species)
) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::scale_y_log10() +
  ggplot2::coord_cartesian(xlim = c(0, 24)) +
  ggplot2::labs(x = "Time (h)", y = "Blood phage (PFU/g)", colour = "Species") +
  ggplot2::theme_bw()
```

## Human dose projection

Two quantitative claims are made about the human simulations, and both are
reproduced here.

* "Monte Carlo simulations predicted rapid elimination of phage in humans,
  resulting in phage blood concentrations being lower than `10^2` PFU/mL (limit
  of quantification by plaque assay) by 12 h."
* "at a maximum dose of `10^12` PFU, phage concentrations are expected to be
  approximately `10^7` PFU/g."

```{r humandose}
h12 <- simPhage(human, 70, 1e12, 0, c(seq(0, 4, by = 0.02), seq(4.1, 24, by = 0.1)))
h12_pfu <- pfu_per_g(h12$Cc, 1e12)

at1h  <- h12_pfu[which.min(abs(h12$time - 1))]
at12h <- h12_pfu[which.min(abs(h12$time - 12))]
signif(c(`1 h` = at1h, `12 h` = at12h), 3)

stopifnot(
  # "approximately 1e7 PFU/g" at the maximum 1e12 PFU dose -- order of magnitude
  at1h > 1e6 && at1h < 1e8,
  # below the 1e2 PFU/mL plaque-assay limit of quantification by 12 h
  at12h < 1e2
)
```

The paper also reports that, because of the short projected human half-life,
"continuous infusion reached steady state levels within 3 hours". A continuous
infusion of the same daily amount reaches its plateau on that timescale here
too.

```{r infusion, fig.cap = "Continuous intravenous infusion in a 70 kg human: approach to steady state."}
inf_times <- seq(0, 12, by = 0.05)
inf_ev <- dplyr::bind_rows(
  data.frame(id = 1L, time = 0, amt = 100, evid = 1L, cmt = "venous",
             dvid = NA_integer_, rate = 100 / 24),
  data.frame(id = 1L, time = inf_times, amt = NA_real_, evid = 0L,
             cmt = NA_character_, dvid = 1L, rate = 0)
)
inf <- as.data.frame(rxode2::rxSolve(
  rxode2::zeroRe(human), inf_ev,
  params = c(WT = 70, DOSE_PHAGE_PFU = 1e12, STUDY_PAML31 = 0),
  maxsteps = 1e6, atol = 1e-14, rtol = 1e-10
))

ggplot2::ggplot(inf, ggplot2::aes(time, pfu_per_g(Cc, 1e12))) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::geom_vline(xintercept = 3, linetype = "dashed") +
  ggplot2::labs(x = "Time (h)", y = "Blood phage (PFU/g)") +
  ggplot2::theme_bw()

css <- inf$Cc[which.min(abs(inf$time - 12))]
frac_at_3h <- inf$Cc[which.min(abs(inf$time - 3))] / css
frac_at_3h

stopifnot(frac_at_3h > 0.95)
```

# Assumptions, deviations and errata

## The fitted observable was co-generated with a twin free-`125I` model that is not reproducible

**This is the most important caveat and it conditions everything else.** The
measured quantity was `125I` scintillation counts, which is labelled phage
*plus* free label. Methods states:

> "A twin PBPK model was constructed to describe free `125I` produced through
> either thermal deconjugation of phage (a relatively slow process) or through
> degradation by the MPS (a relatively fast process)."

That twin model is load-bearing for the fitted observable, but it is described
only in prose. Its equations, compartment list and parameter table appear in no
source:

* The supplement's Equations section covers the phage model only. A full-text
  scan of the supplement returns **zero** occurrences of "iodin", "thyroid",
  "deconjug", "twin" or "125".
* The global sensitivity analysis (Table S4) nevertheless *exercises* two of
  the twin model's parameters -- `T_GI,Lab` and `T_Urine,Lab`, free-label
  transit -- which appear in no parameter table and have no reported values.
  Table S4 lists 14 parameters against Table 1's 10; the other two extras
  (`F_PQ`, `F_KP`) are perturbation factors used only by the sensitivity
  analysis and are not states or estimated parameters of either model.

These model files therefore implement the **phage-disposition system alone**.
The parameter set they carry was estimated jointly with the missing twin model,
so it is conditional on it. Practically, this is why the reconstruction cannot
reproduce the shape of the Figure 2 post-hoc fits: those fitted lines are
**non-monotone** in blood, lung, muscle, bone and stomach, dipping to a nadir
around 8-12 h and then rising and plateauing out to 72 h. The printed phage
system contains no slowly-releasing species that can produce a rise, and no
choice of its parameters produces one. A free-label pool that accumulates as
phage is catabolised does.

This vignette therefore validates against the paper's reported **model-derived
summary statistics** (terminal half-life, mean residence times, stomach-contents
recovery, and the human dose projections) and deliberately does **not** attempt
to replicate Figure 2.

## Table 1 is the final parameter set; the abstract and Results narrative are stale

The abstract, the Results narrative and Table 1 report mutually inconsistent
values for six parameters, and the abstract additionally *swaps* which organ
group gets the small partition coefficient:

| Parameter | Abstract | Results text | **Table 1 (used here)** |
|---|---|---|---|
| K~p,LunKid~ | 0.000138 | -- | **0.395** |
| K~p,LvrSpnSto~ | 0.627 | -- | **2.12e-4** |
| K~p,Crs~ | 0.220 | -- | **0.934** |
| P~S~ | -- | 0.0227 | **4.27e-2** |
| CL~Active~ | -- | 0.0145 | **1.29e-2** |
| T~MPS,deg~ | -- | 0.0301 h | **4.37e-2 h** |

Table 1 was adopted because it, and not the narrative set, reproduces the
paper's own reported outputs: the blood terminal half-life and the liver mean
residence time, both checked above. The narrative set gives a blood half-life
of roughly 3 h against a reported 10.3-12.0 h.

## The PAML-31-1 arm does not reconcile

Table 1 reports `CL_Active` = 1.29e-2, a covariate-adjusted
`CL_Active,PAML` = 1.34e-3, and `theta_PAML31` = -1.94. No standard transform
connects the three: `exp(-1.94)` = 1.85e-3, not 1.34e-3, and Table 1's own two
clearance rows are 9.63-fold apart while the prose claims "10.7-fold lower".
(Note that 0.0145 / 10.7 = 1.36e-3 -- the *narrative's* `CL_Active` is what
reproduces the 10.7-fold claim, which is further evidence that the narrative
and Table 1 come from different runs.)

These models take both clearances **directly from Table 1** and never evaluate
`theta`. That reproduces the *direction* of the PAML-31-1 effect (higher
exposure) but not its magnitude, and it badly overshoots the terminal half-life:
a 9.6-fold lower clearance with an unchanged distribution volume implies a
terminal half-life near 90 h, whereas the paper reports 12.0 h for PAML-31-1 --
essentially the same as the other two phages, alongside a 5.5-fold higher AUC.
A 5.5-fold exposure increase at an unchanged half-life requires a proportional
reduction in volume as well as clearance, which this model structure does not
have. Treat the PAML-31-1 arm as **not reproduced**.

## Blood topology reconstructed from Figure 1

The supplement's venous equation does not conserve mass -- it debits
`Q_Lun * C_Lun,v` where mass balance requires `Q_Lun * C_V` -- and every organ
equation is driven by an arterial concentration `C_a(t)` for which no equation
is ever given. Figure 1 shows unambiguous venous and arterial columns, so the
two-pool `venous -> lung -> arterial -> organs -> venous` topology was
reconstructed from it. The choice is not decision-changing: a single-pool blood
model moves every output reported here by less than 3%.

The venous/arterial volume split (`f_venous` = 0.8) is **not reported by
Echterhof 2026**. It is the convention of the nanoparticle-PBPK framework
lineage this model is built on (Table S3 citations 6-7; the same 0.8 appears in
`Kutumova_2024_albuminNanoparticles_mouse_pbpk`). It is exposed as a `fixed()`
parameter so it can be changed.

## The GI-elimination equation is not printed

Figure 1 draws a "GI Elimination" box fed by active clearance from the liver,
stomach, small intestine and large intestine, and Table 1 estimates a gastric
transit half-life for it, but the supplement prints **no equation** for that
compartment -- only the urine analogue in the Kidney block. The `gut_lumen`
state here is reconstructed by direct analogy with the printed urine equation.
The reconstruction recovers the reported 18.0 %ID stomach-contents recovery to
within the gate applied above, which is evidence the analogy is the right one.

## The brain carries no MPS pool

Figure 1 draws an MPS box for the brain, but the supplement's brain block
prints only the vascular and interstitial equations, with no uptake, release or
degradation terms. The printed equations are followed. Table S3's brain
phagocyte density `M_Brn` is consequently unused and is not declared in the
model files.

## The MPS pathway is nearly inert in the mouse, by construction

The MPS carrying capacity is a fixed absolute number of phage particles
(`10^3.81` PFU per 100,000 phagocytes). At the mouse doses actually
administered (`10^10.1` to `10^11.9` PFU) the whole-body capacity works out to
roughly `5e5` PFU, about `0.001%` of the dose, so the MPS pools saturate
essentially instantly and contribute almost nothing to mouse clearance --
active surface clearance does nearly all of it.

That is not an artefact of this implementation; it is a property of the
published parameter set, and the paper's own global sensitivity analysis
independently confirms it: `CL_Active` has a first-order Sobol index of 0.907
while `A_MPS`, `T_MPS,up` and `T_MPS,deg` are all within noise of zero
(Table S4). In the rat and human, where tabulated phagocyte densities are one
to two orders of magnitude higher, the same capacity relation makes the MPS the
*dominant* route -- which is what produces the very short projected human
half-life, and in turn the paper's own finding of rapid human elimination.

```{r mpsinert}
mps_pool <- max(
  luz24$int_liver + luz24$int_spleen + luz24$int_lung +
    luz24$int_muscle + luz24$int_remainder
)
signif(mps_pool, 3)   # %ID held in the MPS at its peak

stopifnot(mps_pool < 0.01)
```

## `A_MPS` is converted from PFU to %ID by the administered dose

Table 1 gives `A_MPS` in `log10(PFU / 1e5 cells)` while the main text describes
the same quantity as `%ID` per `1e5` phagocytes. The two are reconciled by the
dose: the Interspecies scaling section states that "to simulate PFU per
milliliter, doses were parameterized in PFU, and the **untransformed**
macrophage capacity term was utilized", which implies the %ID-scale fit used a
*transformed* one. These models therefore compute
`A_RES [%ID/1e5 cells] = 10^A_MPS * 100 / DOSE_PHAGE_PFU`.

Three independent checks confirm this reading over the alternative (treating
`10^3.81` as being already in %ID): it reproduces the reported blood terminal
half-life, it reproduces the liver mean residence time, and it reproduces the
Table S4 sensitivity ranking. The alternative reading makes the MPS an
effectively unbounded sink, giving a blood half-life of minutes.

## Numerical note: do not clamp the uptake rate

The saturable uptake rate `k_up,max * (1 - A_res/A_max)` must be coded exactly
as printed. Wrapping it in `max(0, .)` puts a derivative kink precisely at the
operating point `A_res ~ A_max`, which stalls the implicit solver -- with the
clamp in place the PAML-31-1 arm fails to integrate at all. Unclamped, the
relation is self-correcting and every arm solves in a couple of seconds.

## Simulated plasma versus measured whole blood

The model's blood space is plasma: Table S3 calls the fraction `F_V_Pla`, its
mouse value (0.029) is a plasma-like rather than whole-blood fraction, and the
clinical simulation section states that phage was assumed not to distribute
into the red-blood-cell space. The observed `%ID/g`, by contrast, was counted in
whole blood. A constant plasma-to-whole-blood factor of roughly `(1 - Hct)`
between the two is the most likely reason the simulated blood AUC here runs
above the reported whole-blood NCA AUC, and it is consistent with the simulated
1-minute concentration (28 %ID/g) against the reported peak of about 15 %ID/g.
This is offered as the likely explanation, not as an established fact -- the
paper does not state a haematocrit correction. Note that the quantities gated
on above (terminal half-life, mean residence times, the steady-state fraction)
are all invariant to a constant scale factor, which is why they were chosen.

## Other defects noted in the source

* Figure 2's legend says **LUZ19**; the abstract, Methods, Results and Table S1
  all say **Luz24**.
* Table S3's row `F_V_Spn` is captioned "Fraction of BW as **Lung**" -- a
  copy-paste label error; the row is the spleen.
* The supplement's small-intestine block closes with two equations named for the
  **spleen** (`k_Spn,up`, `A_Spn,max`); they are plainly meant to be the small
  intestine.
* The skin and carcass blocks each define `dC_x,V/dt` twice, the second being
  the interstitial equation mislabelled.

## Naming

The per-organ phagocyte pool uses the already-registered `int_<organ>`
sub-compartment namespace, whose founding example
(`Kutumova_2024_albuminNanoparticles_mouse_pbpk`) uses it for exactly this
state -- the internalised, phagocytosed nanoparticle pool of a whole-body
mononuclear-phagocyte-system PBPK. The `stomach` organ was added to the
`pbpkSubCompartmentRegex` organ vocabulary, and `vp_stomach` registered, both of
which the compartment register calls routine extensions. The GI-elimination
state reuses the registered `gut_lumen` canonical.
