---
title: "N,N-dimethyltryptamine (Eckernas 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{N,N-dimethyltryptamine (Eckernas 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Eckernas_2022_dimethyltryptamine"))$reference`
* Description: `r rxode2::rxode(readModelDb("Eckernas_2022_dimethyltryptamine"))$description`
* Article: <https://doi.org/10.1111/cts.13410> (open access, PMC9747126)

The paper builds one model in three sequential steps: a two-compartment
population PK model for plasma DMT, extended with its major metabolite
indole-3-acetic acid (IAA), and finally an effect-compartment sigmoidal Emax
model for the subjective intensity of the psychedelic experience, fitted with
the PK parameters held at their population estimates ("population PK
parameters and data" approach). All parameters are in the single Table 1, so
the package carries the whole cascade as one model file.

The same authors' later EEG analysis reuses this PK model with its parameters
fixed; those three models ship as `Eckernas_2023_dimethyltryptamine_alphaPower`,
`_betaPower` and `_LZc`.

# Population

Thirteen healthy volunteers (7 men, 6 women; median age 33 years, range 22-48)
took part in a placebo-controlled, fixed-order study at the NIHR Imperial
Clinical Research Facility, London. Each received placebo at the first visit
and, one week later, a single intravenous bolus of DMT fumarate: 7 mg (n = 3),
14 mg (n = 4), 18 mg (n = 1) or 20 mg (n = 5). Nine plasma samples per subject
were taken up to 60 min after dosing, and the subject rated the intensity of
the experience on a 0-10 scale (0 = no effect, 10 = the most intense experience
imaginable) every minute for the first 20 min. The analysis used 93 DMT and 87
IAA concentrations and 273 intensity ratings (Results, first paragraph).
Placebo-visit data showed no measurable DMT and no subjective effect and were
not modelled. The paper does not tabulate body weight or race.

The same information is available programmatically via
`readModelDb("Eckernas_2022_dimethyltryptamine")()$population`.

# Source trace

Every `ini()` value carries an in-file comment naming its source. The table
collects them.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` (DMT CL) | 26.0 L/min | Table 1 |
| `lvc` (DMT Vc) | 221 L | Table 1 |
| `lq` (DMT Q) | 2.99 L/min | Table 1 |
| `lvp` (DMT Vp) | 59.0 L | Table 1 |
| `fm` (fraction of DMT elimination forming IAA) | 1 (fixed) | Methods, Pharmacokinetic model development |
| `lcl_iaa` (CL(m)) | 0.093 L/min | Table 1 |
| `lvc_iaa` (V(m)) | 9.55 L | Table 1 |
| `emax` | 10 (fixed) | Table 1; Results, PK/PD model |
| `lec50` (EC50,e) | 94.7 nM | Table 1 |
| `lke0` | 1.38 1/min | Table 1 |
| `lhill` (gamma) | 2.87 | Table 1 |
| `etalcl` | 0.473^2 = 0.2237 | Table 1, BSV CL 47.3 %CV |
| `etalvc_iaa` | 0.290^2 = 0.0841 | Table 1, BSV V(m) 29.0 %CV |
| `etalec50` | 0.386^2 = 0.1490 | Table 1, BSV EC50,e 38.6 %CV |
| `etalhill` | 0.773^2 = 0.5975 | Table 1, BSV gamma 77.3 %CV |
| `propSd` (DMT) | 0.500 | Table 1, residual error DMT 50.0 %CV |
| `propSd_iaa` | 0.190 | Table 1, residual error IAA 19.0 %CV |
| `addSd_psychedelicintensity` | 0.82 (logit scale) | Table 1; Equation 1 |
| `d/dt(central)`, `d/dt(peripheral1)` | two-compartment, first-order elimination | Methods; Results, Population PK model |
| `d/dt(central_iaa) = fm * kel * central - kel_iaa * central_iaa` | n/a | Methods: IAA formation rate equals the DMT elimination rate |
| `d/dt(effect) = ke0 * (Cc - effect)` | n/a | Methods, effect-compartment equation |
| `psychedelicintensity = emax * Ce^gamma / (EC50^gamma + Ce^gamma)` | E0 = 0 | Methods, Emax equation; Results: all subjects had a baseline of zero |
| `psychedelicintensity ~ logitNorm(sd, 0, 10)` | n/a | Equation 1: `y = 10 * exp(ln(lambda/(10 - lambda)) + eps) / (1 + exp(...))` |

# Dose units

The model runs in molar units: doses are nmol of DMT base and concentrations
are nmol/L (= nM), the unit of EC50,e. The paper states doses as mg of DMT
fumarate and that they "were converted and expressed as nanomoles of DMT base",
without printing the factor. The vignette uses the 1:1 salt stoichiometry,
`DMT (C12H16N2, 188.27 g/mol) + fumaric acid (C4H4O4, 116.07 g/mol)`, giving
304.34 g/mol, or 3286 nmol DMT base per mg of salt.

```{r dose-conversion}
mw_dmt      <- 188.27
mw_fumarate <- 116.07
nmolPerMg   <- 1e6 / (mw_dmt + mw_fumarate)
doseMg      <- c(1, 4, 7, 14, 20)  # Figure 4 simulation doses (mg DMT fumarate)
tibble(
  `Dose (mg DMT fumarate)` = doseMg,
  `Dose (nmol DMT base)`   = round(doseMg * nmolPerMg),
  `C0 = Dose / Vc (nM)`    = round(doseMg * nmolPerMg / 221, 1)
) |>
  knitr::kable(caption = "Molar dose conversion used throughout this vignette.")
```

# Virtual cohort

The paper's Figure 4 simulated 100 subjects at each of five IV bolus doses. The
vignette uses 200 subjects per dose, the most it permits, so that the
dose-response statistics checked below sit well clear of Monte-Carlo noise. No
covariates enter the model.

```{r cohort}
# rxSetSeed() fixes rxode2's draw for a given solver thread count only, so the
# assertions below are written to hold for any cohort the model can produce.
rxode2::rxSetSeed(2022)
set.seed(2022)

nPerDose <- 200L
obsGrid  <- seq(0, 60, by = 0.25)

make_cohort <- function(n, mg, id_offset) {
  ids <- id_offset + seq_len(n)
  dose <- tibble(id = ids, time = 0, amt = mg * nmolPerMg, evid = 1L,
                 cmt = "central", dvid = NA_integer_)
  obs  <- tidyr::expand_grid(id = ids, time = obsGrid) |>
    mutate(amt = 0, evid = 0L, cmt = "central", dvid = 1L)
  bind_rows(dose, obs) |>
    mutate(dose_mg = mg) |>
    arrange(id, time, desc(evid))
}

events <- bind_rows(lapply(seq_along(doseMg), function(i) {
  make_cohort(nPerDose, doseMg[i], id_offset = (i - 1L) * nPerDose)
}))
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

Observation rows are placed on the `central` state with `dvid = 1`; rxode2
returns every model variable (`Cc`, `Cc_iaa`, `psychedelicintensity`) on each
row. Residual error is added afterwards in R, following the paper's
definitions, so each endpoint gets its own draw.

```{r simulate}
mod <- readModelDb("Eckernas_2022_dimethyltryptamine")
sim <- rxode2::rxSolve(mod, events = events, keep = "dose_mg",
                       returnType = "data.frame") |>
  mutate(
    dose = factor(paste(dose_mg, "mg"), levels = paste(doseMg, "mg")),
    # Proportional residual error on DMT and IAA (Table 1).
    Cc_obs     = Cc * (1 + 0.500 * rnorm(n())),
    Cc_iaa_obs = Cc_iaa * (1 + 0.190 * rnorm(n())),
    # Equation 1: additive error on the logit scale bounded by 0 and 10.
    rating_obs = 10 * plogis(qlogis(psychedelicintensity / 10) + 0.82 * rnorm(n()))
  )
stopifnot(!anyNA(sim$Cc), !anyNA(sim$Cc_iaa), !anyNA(sim$psychedelicintensity))
```

# Replicate published figures

## Figure 1 -- DMT and IAA concentrations by dose

Observed medians in Figure 1b-c are drawn as approximate points read off the
published figure by the maintainers. The study's 18 mg and 20 mg subjects
were pooled in that figure; the simulation shows 20 mg.

```{r figure-1}
fig1_obs <- tribble(
  ~analyte, ~dose,   ~time, ~conc,
  "IAA",    "7 mg",   5,    1200,
  "IAA",    "7 mg",  26,    1900,
  "IAA",    "7 mg",  44,    1550,
  "IAA",    "7 mg",  60,    1550,
  "IAA",    "14 mg",  5,    2000,
  "IAA",    "14 mg", 20,    3100,
  "IAA",    "14 mg", 40,    3200,
  "IAA",    "14 mg", 60,    3050,
  "IAA",    "20 mg",  3,    2900,
  "IAA",    "20 mg", 20,    4700,
  "IAA",    "20 mg", 38,    4750,
  "IAA",    "20 mg", 60,    4300,
  "DMT",    "7 mg",   2,      45,
  "DMT",    "7 mg",   5,      30,
  "DMT",    "14 mg",  3,     105,
  "DMT",    "14 mg",  5,     100,
  "DMT",    "20 mg",  2,     135,
  "DMT",    "20 mg",  5,     130,
  "DMT",    "20 mg", 20,      18
) |>
  mutate(dose = factor(dose, levels = paste(doseMg, "mg")))

sim |>
  filter(dose_mg %in% c(7, 14, 20), time > 0) |>
  select(dose, time, DMT = Cc_obs, IAA = Cc_iaa_obs) |>
  pivot_longer(c(DMT, IAA), names_to = "analyte", values_to = "conc") |>
  group_by(analyte, dose, time) |>
  summarise(Q05 = quantile(conc, 0.05), Q50 = median(conc),
            Q95 = quantile(conc, 0.95), .groups = "drop") |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue4") +
  geom_point(data = fig1_obs, aes(time, conc), colour = "black") +
  facet_grid(analyte ~ dose, scales = "free_y") +
  labs(x = "Time (min)", y = "Concentration (nM)",
       title = "Figure 1 -- simulated DMT and IAA (median, 5th-95th percentile)",
       caption = "Replicates Figure 1b-c of Eckernas 2022. Points: approximate observed medians read from the figure.")
```

The simulated IAA medians follow the observed medians at all three dose
levels. Once DMT has been eliminated, the IAA level is set mainly by the molar
dose and `V(m)`, so this agreement supports the dose conversion. The early DMT medians lie somewhat above the
observed values at 2-5 min. Those observed values come from three to six
subjects per panel, and the 18/20 mg panel includes the one 18 mg subject.

## Figure 2 -- intensity rating versus time

```{r figure-2}
fig2_obs <- tribble(
  ~dose,   ~time, ~rating,
  "7 mg",    2.5,  4.5,
  "7 mg",    7,    5,
  "7 mg",   15.5,  2,
  "14 mg",   2.5,  7,
  "14 mg",   7,    4,
  "14 mg",  15.5,  1,
  "20 mg",   2.5,  9.5,
  "20 mg",   7,    8,
  "20 mg",  15.5,  1
) |>
  mutate(dose = factor(dose, levels = paste(doseMg, "mg")))

sim |>
  filter(dose_mg %in% c(7, 14, 20), time <= 20) |>
  group_by(dose, time) |>
  summarise(Q05 = quantile(rating_obs, 0.05), Q50 = median(rating_obs),
            Q95 = quantile(rating_obs, 0.95), .groups = "drop") |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "firebrick") +
  geom_line(colour = "firebrick4") +
  geom_point(data = fig2_obs, aes(time, rating), colour = "black") +
  facet_wrap(~dose) +
  coord_cartesian(ylim = c(0, 10)) +
  labs(x = "Time (min)", y = "Subjective intensity rating",
       title = "Figure 2 -- simulated intensity rating (median, 5th-95th percentile)",
       caption = "Replicates Figure 2b of Eckernas 2022. Points: approximate observed medians read from the figure.")
```

The 14 mg and 20 mg simulated medians track the observed medians. At 7 mg
the observed medians come from only three subjects and lie above the simulated
median after the peak; they fall inside the simulated 5th-95th percentile band,
as they do inside the paper's own prediction interval for that panel.

## Figure 4 -- maximum achieved intensity by dose

Figure 4 shows the distribution of each simulated subject's maximum rating over
the 20-min rating window. The vignette takes the maximum of the individual
prediction (no residual error), which reproduces the published boxes; the paper
does not say whether residual error was included.

```{r figure-4}
maxRating <- sim |>
  filter(time <= 20) |>
  group_by(dose, dose_mg, id) |>
  summarise(max_rating = max(psychedelicintensity), .groups = "drop")

# Figure 4 whiskers span the 5th-95th percentiles.
maxRating |>
  group_by(dose) |>
  summarise(
    ymin   = quantile(max_rating, 0.05),
    lower  = quantile(max_rating, 0.25),
    middle = median(max_rating),
    upper  = quantile(max_rating, 0.75),
    ymax   = quantile(max_rating, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(dose, ymin = ymin, lower = lower, middle = middle,
             upper = upper, ymax = ymax)) +
  geom_boxplot(stat = "identity") +
  labs(x = "Dose (mg DMT fumarate)", y = "Maximum subjective intensity rating",
       title = "Figure 4 -- maximum achieved intensity",
       caption = "Replicates Figure 4 of Eckernas 2022.")
```

```{r figure-4-check}
# Medians of the Figure 4 boxes (read from the figure) and the Results text:
# "A median maximum effect of 0, 2, 4, 8, and 9 ... The proportion of the
# population achieving a maximum rating above five increased from zero at the
# lowest dose to 4, 42, 92, and 100%".
published4 <- tibble(
  dose_mg        = doseMg,
  median_figure  = c(0, 1.35, 3.95, 7.95, 9.2),
  median_text    = c(0, 2, 4, 8, 9),
  pct_above5     = c(0, 4, 42, 92, 100)
)
fig4 <- maxRating |>
  group_by(dose_mg) |>
  summarise(sim_median = median(max_rating),
            sim_pct_above5 = 100 * mean(max_rating > 5), .groups = "drop") |>
  left_join(published4, by = "dose_mg")

fig4 |>
  mutate(across(where(is.double), ~ round(.x, 2))) |>
  rename(
    "Dose (mg)"                  = dose_mg,
    "Simulated median max"       = sim_median,
    "Simulated % max > 5"        = sim_pct_above5,
    "Figure 4 median"            = median_figure,
    "Text median"                = median_text,
    "Published % max > 5"        = pct_above5
  ) |>
  knitr::kable(caption = "Maximum rating by dose: simulated versus Eckernas 2022.")

# A 3000-subject-per-dose run of this model gives medians 0.02 / 1.35 / 4.12 /
# 7.90 / 9.10 and percentages above 5 of 0 / 3.0 / 34.8 / 91.1 / 98.7. With 200
# subjects per dose the standard error of a median is about 0.2 rating units and
# of a percentage at most 3.5 points, so the bounds below are ~4 standard
# errors wide. The 7 mg percentage sits furthest from the paper (model 35%
# versus 42%); one 200-subject draw realised 27.5%, inside the bound. A two-fold
# error in EC50, a dose or a unit moves the 7 mg and 14 mg medians by several
# units and fails them.
stopifnot(
  all(abs(fig4$sim_median - fig4$median_figure) < 0.8),
  all(abs(fig4$sim_pct_above5 - fig4$pct_above5) < 20)
)
```

The simulated medians agree with the Figure 4 boxes. The Results text rounds
the 4 mg median to 2, whereas the figure box and the model sit at about 1.4.
The model's share of 7 mg subjects above a rating of 5 is about 35%, against
42% in the text; the paper's own 100-subject simulation carries a standard
error of about 5 points on that figure.

# PKNCA validation

The paper reports no NCA table. The check below runs PKNCA on a typical-value
solve and compares against quantities the model implies in closed form: DMT
`AUC0-inf = Dose / CL`, IAA `AUC0-inf = fm * Dose / CL(m)` (independent of
the DMT parameters because every mole of DMT becomes IAA), the DMT terminal
half-life from the two-compartment eigenvalue, and the IAA half-life
`ln(2) * V(m) / CL(m)`, which is the slowest rate in the system.

```{r pknca}
ncaDoseMg <- c(7, 14, 20)
ncaGrid <- sort(unique(c(seq(0, 10, by = 0.1), seq(10, 720, by = 2))))
ncaEvents <- bind_rows(lapply(seq_along(ncaDoseMg), function(i) {
  bind_rows(
    tibble(id = i, time = 0, amt = ncaDoseMg[i] * nmolPerMg, evid = 1L,
           cmt = "central", dvid = NA_integer_),
    tibble(id = i, time = ncaGrid, amt = 0, evid = 0L, cmt = "central", dvid = 1L)
  ) |>
    mutate(treatment = paste(ncaDoseMg[i], "mg"))
}))

# Typical values: omega = NA and sigma = NA drop all random effects.
typ <- rxode2::rxSolve(mod, events = ncaEvents, omega = NA, sigma = NA,
                       keep = "treatment", rtol = 1e-10, atol = 1e-12,
                       returnType = "data.frame")

concLong <- bind_rows(
  # DMT falls ~13 terminal half-lives by 240 min; later points are solver noise.
  typ |> filter(time <= 240) |> transmute(id, time, treatment, analyte = "DMT", Cc),
  typ |> transmute(id, time, treatment, analyte = "IAA", Cc = Cc_iaa)
) |>
  filter(!is.na(Cc))
stopifnot(all(concLong$Cc >= -1e-6 * max(concLong$Cc)))
concLong$Cc <- pmax(concLong$Cc, 0)

conc_obj <- PKNCA::PKNCAconc(concLong, Cc ~ time | analyte + treatment + id)
dose_obj <- PKNCA::PKNCAdose(
  ncaEvents |> filter(evid == 1) |> select(id, time, amt, treatment),
  amt ~ time | treatment + id
)
intervals <- data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                        aucinf.obs = TRUE, half.life = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r nca-compare}
th <- list(cl = 26.0, vc = 221, q = 2.99, vp = 59.0, cl_iaa = 0.093, vc_iaa = 9.55)
kel <- th$cl / th$vc
k12 <- th$q / th$vc
k21 <- th$q / th$vp
beta <- ((kel + k12 + k21) - sqrt((kel + k12 + k21)^2 - 4 * kel * k21)) / 2

reference <- tidyr::expand_grid(analyte = c("DMT", "IAA"), dmg = ncaDoseMg) |>
  mutate(
    treatment  = paste(dmg, "mg"),
    doseNmol   = dmg * nmolPerMg,
    aucinf.obs = ifelse(analyte == "DMT", doseNmol / th$cl, doseNmol / th$cl_iaa),
    half.life  = ifelse(analyte == "DMT", log(2) / beta, log(2) * th$vc_iaa / th$cl_iaa),
    cmax       = ifelse(analyte == "DMT", doseNmol / th$vc, NA_real_)
  ) |>
  select(analyte, treatment, cmax, aucinf.obs, half.life)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = reference,
  by        = c("analyte", "treatment"),
  params    = c("cmax", "aucinf.obs", "half.life"),
  units     = c(cmax = "nM", aucinf.obs = "nM*min", half.life = "min"),
  tolerance_pct = 5
)
knitr::kable(cmp, caption = "PKNCA on the typical-value solve versus the model's closed-form values. * differs by >5%.")

# Deterministic solve against its own closed form: the residual is numerical
# (trapezoid and terminal-slope fit), measured below 1%.
pp <- as.data.frame(nca_res) |>
  filter(PPTESTCD %in% c("aucinf.obs", "half.life")) |>
  left_join(reference |> pivot_longer(c(aucinf.obs, half.life),
                                      names_to = "PPTESTCD", values_to = "ref") |>
              select(analyte, treatment, PPTESTCD, ref),
            by = c("analyte", "treatment", "PPTESTCD"))
stopifnot(nrow(pp) == 12L, all(abs(pp$PPORRES / pp$ref - 1) < 0.02))
```

The DMT `Cmax` row is the time-zero concentration of the bolus, `Dose / Vc`.
IAA has no closed-form `Cmax` and is omitted from that row.

# Assumptions and deviations

- **Molar dose conversion.** The paper converts DMT fumarate doses to nmol of
  DMT base without printing the factor. The vignette uses 1:1 salt
  stoichiometry (304.34 g/mol, 3286 nmol/mg). The sibling
  `Eckernas_2023_dimethyltryptamine_EEG` article infers about 3150 nmol/mg from
  a figure axis, 4% lower. The Figure 4 medians reproduced here, which depend on
  the conversion through EC50,e, support the stoichiometric value.
- **Variance scale of the BSV terms.** Table 1 reports BSV as %CV without
  stating the conversion. The variances are `(CV/100)^2`, because the same
  authors' 2023 analysis carries BSV on CL forward as `$OMEGA 0.224 FIX`, which
  is `0.473^2` rather than `log(1 + 0.473^2) = 0.201`. The other three BSV
  terms were converted the same way.
- **PK parameters are the paper's estimates.** In the final PK/PD step the
  population PK parameters were held at their PK-step estimates. The library
  model presents all of Table 1 as one model, so the PK values are the
  estimated values and are not wrapped in `fixed()`; `Emax = 10` and the
  metabolic fraction of 1, which the paper held constant, are `fixed()`.
- **IAA baseline.** IAA was assayed as the change from the endogenous
  baseline, so `Cc_iaa` is the DMT-derived increment and starts at zero.
- **Correlation between EC50,e and gamma.** The paper estimated a 93%
  correlation between those two random effects but dropped it from the final
  model for ill-conditioning. The model follows the final, uncorrelated form.
- **Rating at zero concentration.** With a zero baseline the prediction is 0
  before dosing, where the logit transform in Equation 1 is undefined. The paper
  notes that the transform only approaches the scale limits asymptotically. In
  this vignette the residual draw at a zero prediction returns 0.
- **Figure 4 maximum.** The maximum rating is taken from the individual
  prediction without residual error, over a 0.25-min grid from 0 to 20 min.
- **Observed medians** overlaid on Figures 1 and 2 are approximate values read
  from the published figures by the maintainers and are not used in any
  assertion.
- No erratum for this article was found on the publisher page or in Europe PMC
  as of 2026-10-10.
