---
title: "Obeticholic acid (Edwards 2016)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Obeticholic acid (Edwards 2016)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Edwards JE, LaCerte C, Peyret T, Gosselin NH, Marier JF,
  Hofmann AF, Shapiro D. Modeling and Experimental Studies of Obeticholic
  Acid Exposure and the Impact of Cirrhosis Stage. *Clin Transl Sci*
  2016;9(6):328-336. doi:10.1111/cts.12421
* Article: <https://doi.org/10.1111/cts.12421>

Obeticholic acid (OCA) is a semisynthetic bile acid and selective farnesoid
X receptor (FXR) agonist. The packaged model in
`inst/modeldb/specificDrugs/Edwards_2016_obeticholicAcid_pbpk.R` is the
semi-mechanistic physiologic pharmacokinetic model of Edwards et al. It is
an OCA-specific adaptation of the Molino 1986 chenodeoxycholic acid model:
the enteral system is collapsed to a single gut space, endogenous synthesis
is removed (OCA is exogenous), and bacterial 7-alpha-dehydroxylation is
omitted because the 6-alpha-ethyl group sterically blocks it.

Three chemical forms circulate: unconjugated OCA and its two hepatic
conjugates, glyco-OCA and tauro-OCA. Each form occupies its own copy of the
anatomical spaces, because blood flows are independent of chemical structure
while transport and biotransformation rates are not. The spaces are systemic
circulation, portal circulation, sinusoidal circulation, liver, bile duct,
gallbladder and gut.

`r ""`Unconjugated OCA has no bile duct or gallbladder state: Figure 1 shows
no liver-to-bile-duct arrow in the OCA column (there is no `t20` in
Supplemental Table 3), so only the conjugates are secreted into bile. That
gives 5 + 7 + 7 = 19 ODE states plus an oral depot.

Two classes of flux coexist and are dimensionally different:

* **Flows** `f` (L/h) act on the **concentration** of the source space, so
  they need that space's volume.
* **Transport** `t`, **biotransformation** `b`, `Ka` and the gallbladder
  output `f23` (1/h) act directly on the **amount** in the source state.

```{r load-model}
mod <- modellib("Edwards_2016_obeticholicAcid_pbpk")
mod
```

# Population

```{r population, echo = FALSE}
knitr::kable(
  tibble::tribble(
    ~Item,                ~Value,
    "Species",            "Human",
    "Subjects",           "399 across 5 phase 1/2 studies",
    "Development data",   "Study 1 healthy volunteers (n = 160; 8,248 samples); Study 2 hepatic impairment (n = 32; 928 samples)",
    "Validation data",    "Studies 3 (n = 24) and 4 (n = 160) healthy; Study 5 / PESTO cirrhosis with portal hypertension (n = 23)",
    "Age",                "Healthy mean (SD) 37.0 (9.8) y; hepatic impairment 55.0 (5.6) y",
    "Weight",             "Healthy mean (SD) 76.4 (11.8) kg; hepatic impairment 81.7 (16.9) kg",
    "Sex",                "59% male (healthy pool), 72% male (hepatic impairment cohort)",
    "Doses",              "5, 10 and 25 mg oral; single dose and once-daily to steady state",
    "Software",           "Phoenix NLME v1.3, Lindstrom-Bates FOCE"
  ),
  caption = "Population summary (Edwards 2016 Results and Supplementary Table S1)."
)
```

Child-Pugh classification, per the paper's Results section, is Class A / mild
= 5-6 points, Class B / moderate = 7-9 points, Class C / severe = 10-15
points. The hepatic-impairment cohort had mean (SD) Child-Pugh score 8.0
(2.0).

# Source trace

Every value in `ini()` and every structural equation, with the location it
came from. "Suppl Table 3/4" refers to the supplementary files distributed
with the article (`CTS-9-328-s004.docx` and `CTS-9-328-s005.docx`).

```{r source-trace, echo = FALSE}
knitr::kable(
  tibble::tribble(
    ~Quantity, ~`Paper name`, ~Value, ~Source,
    "Compartment volumes (7)", "Systemic / Portal / Sinusoidal / Liver / Bile duct / Gallbladder / Intestine", "2.46 / 0.42 / 0.12 / 0.95 / 0.02 / 0.02 / 0.90 L", "Figure 1, Volumes legend inside the panel (p. 330)",
    "Model topology", "f, t, b arrows", "see Figure 1", "Figure 1 diagram (p. 330)",
    "Hepatic portal flow", "f3", "39.6 L/h (fixed)", "Suppl Table 3",
    "Hepatic arterial flow", "f4", "14.4 L/h (fixed)", "Suppl Table 3",
    "Systemic -> portal flow", "f1", "= f2 + f3 (39.6 L/h)", "Figure 1 arrow; value by blood-flow continuity on the portal space",
    "Portal -> systemic shunt", "f2", "= f3(healthy) - f3(impaired); 0 in healthy", "Results, Hepatic impairment (equal-magnitude rule); Figure 1 dotted arrow",
    "Sinusoidal -> systemic flow", "f5", "= f3 + f4", "Figure 1 arrow; value by blood-flow continuity on the sinusoidal space",
    "Bile duct -> gallbladder", "f22", "0.856 L/h", "Suppl Table 3",
    "Bile duct -> gut", "f24", "7.29 L/h", "Suppl Table 3",
    "Gallbladder -> gut", "f23", "1.2 /h (fixed)", "Suppl Table 3",
    "Fecal elimination", "Kout", "0.612 L/h", "Suppl Table 3",
    "Sinusoidal -> liver", "t10 / t9 / t11", "1698 / 1210 / 1615 /h", "Suppl Table 3",
    "Liver -> sinusoidal", "t13 / t12", "1.62 / 1.62 /h (both fixed)", "Suppl Table 3",
    "Liver -> bile duct", "t19 / t21", "7.44 / 9.28 /h", "Suppl Table 3",
    "Gut -> portal", "t34 / t33 / t35", "0.857 / 0.904 / 1.62 /h", "Suppl Table 3",
    "Conjugation", "b15 / b16", "1.44 / 0.312 /h", "Suppl Table 3",
    "Deconjugation", "b36 / b37", "0.0431 / 0.0200 /h", "Suppl Table 3",
    "Oral absorption", "Ka", "5.32 /h", "Suppl Table 3",
    "BSV Ka", "-", "195% CV -> omega^2 = 1.5692", "Suppl Table 3",
    "BSV f22", "-", "78.1% CV -> omega^2 = 0.4762", "Suppl Table 3",
    "Residual error (healthy)", "-", "prop 88.0 / 62.6 / 71.6%; add 0.546 / 0.675 / 0.469 nM", "Suppl Table 3",
    "Hepatic uptake effect", "-", "-0.132 / -1.86 / -2.37", "Suppl Table 4",
    "Tauro-conjugation effect", "-", "0.00481 / 1.05 / 1.56", "Suppl Table 4",
    "f4 multipliers", "-", "1.408 / 1.625 / 1.915", "Suppl Table 2",
    "f3 multipliers", "-", "0.91 / 0.635 / 0.554", "Suppl Table 2",
    "Liver volume multipliers", "-", "0.891 / 0.71 / 0.61", "Suppl Table 2",
    "Gallbladder contraction window", "-", "90 min from meal start", "Results, Source data",
    "Molecular weights", "-", "OCA 420.6, glyco-OCA 477.7, tauro-OCA 527.8 g/mol", "Results, Bioanalytical"
  ),
  caption = "Source trace for the Edwards 2016 obeticholic acid model."
)
```

Three quantities in that table are **derived rather than printed**, and are
flagged as such in the model file. `f1` and `f5` appear as labelled arrows in
Figure 1 but are not tabulated; their values follow from blood-flow
continuity on the portal and sinusoidal spaces given the printed `f3` and
`f4`. `f2` follows from the printed equal-magnitude shunting rule. No value
in this model is taken from outside the on-disk sources.

# Virtual cohort

Study 2 is the hepatic-impairment study: a single 10 mg oral dose with
sampling to 216 h, eight subjects each with normal function and Child-Pugh
A / B / C. We simulate 100 subjects per arm, which is ample for the checks
below and well inside the 200-per-arm cap.

```{r cohort}
MW <- c(oca = 420.6, goca = 477.7, toca = 527.8)  # g/mol, Bioanalytical section
DOSE_MG <- 10
dose_nmol <- DOSE_MG / MW[["oca"]] * 1e6

N_PER_ARM <- 100L
obs_times <- seq(0, 216, by = 0.5)

# Standardized inpatient meals. The paper records meal times per subject but
# does not publish them; see Assumptions and deviations.
meal_starts <- as.vector(outer(c(4, 10), seq(0, 216, by = 24), "+"))
meal_starts <- sort(meal_starts[meal_starts <= 216])
MEAL_DUR <- 1.5  # 90 minutes, Results / Source data

meal_flag <- function(t) {
  as.integer(vapply(t, function(ti) any(ti >= meal_starts & ti < meal_starts + MEAL_DUR), logical(1)))
}

arms <- tibble::tribble(
  ~arm,        ~HEPIMP_MILD, ~HEPIMP_MOD, ~HEPIMP_SEV,
  "Normal",    0,            0,           0,
  "Mild",      1,            0,           0,
  "Moderate",  0,            1,           0,
  "Severe",    0,            0,           1
)

build_events <- function(mild, mod, sev, n) {
  per_id <- data.frame(
    time = c(0, obs_times),
    amt  = c(dose_nmol, rep(NA_real_, length(obs_times))),
    evid = c(1L, rep(0L, length(obs_times))),
    # Observation rows sit on the ODE state and select the endpoint with
    # dvid; never name an algebraic observable in `cmt`.
    cmt  = c("depot", rep("systemic_oca", length(obs_times))),
    dvid = c(NA_integer_, rep(1L, length(obs_times)))
  )
  ev <- per_id[rep(seq_len(nrow(per_id)), times = n), ]
  ev$id <- rep(seq_len(n), each = nrow(per_id))
  ev$HEPIMP_MILD <- mild
  ev$HEPIMP_MOD <- mod
  ev$HEPIMP_SEV <- sev
  ev$MEAL_FLAG <- meal_flag(ev$time)
  ev[order(ev$id, ev$time, -ev$evid), ]
}
```

# Simulation

Two solves are used, for two different purposes.

* A **typical-value** solve (`rxode2::zeroRe()`, `omega = NA`) reproduces the
  Table 1 exposure ratios. This is deterministic: both sides of the
  comparison use the same parameters, so exact bounds are appropriate.
* A **stochastic** solve carries the published BSV on `Ka` and `f22` and
  feeds the NCA section.

```{r simulate-typical}
sim_typical <- function(row) {
  ev <- build_events(row$HEPIMP_MILD, row$HEPIMP_MOD, row$HEPIMP_SEV, 1L)
  s <- rxSolve(zeroRe(mod), ev, omega = NA, sigma = NA, returnType = "data.frame")
  s$arm <- row$arm
  s
}
typ <- do.call(rbind, lapply(split(arms, seq_len(nrow(arms))), sim_typical))

# Total OCA is the molar sum of the three analytes (Results, Source data),
# converted to ng/mL with the published molecular weights.
typ <- typ %>%
  mutate(
    total_sys_ng = (Cc * MW[["oca"]] + Cc_goca * MW[["goca"]] + Cc_toca * MW[["toca"]]) / 1000,
    arm = factor(arm, levels = arms$arm)
  )
```

Liver concentrations are reported by the paper in ng/mL. The model returns
`Cliver` as the molar sum divided by the (impairment-adjusted) liver volume,
so the conversion needs the analyte-weighted molecular weight rather than a
single constant.

```{r liver-conc}
# Functional liver volume by arm: 0.95 L scaled by the Supplemental Table 2
# multipliers for Child-Pugh A / B / C.
v_liv_arm <- c(Normal = 0.95, Mild = 0.95 * 0.891, Moderate = 0.95 * 0.710,
               Severe = 0.95 * 0.610)

typ <- typ %>%
  mutate(
    liver_ng = (liver_oca * MW[["oca"]] + liver_goca * MW[["goca"]] +
                  liver_toca * MW[["toca"]]) / 1000,
    liver_conc_ng = liver_ng / v_liv_arm[as.character(arm)]
  )
```

```{r auc-helper}
trapz <- function(t, y) sum(diff(t) * (head(y, -1) + tail(y, -1)) / 2)

exposure <- typ %>%
  group_by(arm) %>%
  summarise(
    sys_auc   = trapz(time, total_sys_ng),
    sys_cmax  = max(total_sys_ng),
    sys_cavg  = trapz(time[time <= 24], total_sys_ng[time <= 24]) / 24,
    liv_auc   = trapz(time, liver_conc_ng),
    liv_cmax  = max(liver_conc_ng),
    .groups = "drop"
  ) %>%
  mutate(
    sys_auc_ratio  = sys_auc / sys_auc[arm == "Normal"],
    liv_auc_ratio  = liv_auc / liv_auc[arm == "Normal"],
    sys_cmax_ratio = sys_cmax / sys_cmax[arm == "Normal"],
    liv_cmax_ratio = liv_cmax / liv_cmax[arm == "Normal"]
  )
knitr::kable(exposure, digits = 3, caption = "Simulated typical-value exposure by hepatic-impairment arm.")
```

# Replicating Table 1

Table 1 of Edwards 2016 reports systemic and liver exposure ratios relative
to normal hepatic function. This is the paper's central quantitative claim:
systemic exposure rises steeply with cirrhosis stage while liver exposure
rises only marginally.

```{r table1-compare}
published <- tibble::tribble(
  ~arm,       ~pub_sys_auc_ratio, ~pub_liv_auc_ratio,
  "Normal",   1.00,               1.00,
  "Mild",     1.35,               1.12,
  "Moderate", 8.03,               1.47,
  "Severe",   13.2,               1.74
)

cmp <- exposure %>%
  select(arm, sys_auc_ratio, liv_auc_ratio) %>%
  left_join(published, by = "arm") %>%
  mutate(
    sys_pct_diff = 100 * (sys_auc_ratio - pub_sys_auc_ratio) / pub_sys_auc_ratio,
    liv_pct_diff = 100 * (liv_auc_ratio - pub_liv_auc_ratio) / pub_liv_auc_ratio
  )

cmp %>%
  rename(
    "Arm" = arm,
    "Systemic AUC ratio (simulated)" = sys_auc_ratio,
    "Systemic AUC ratio (published)" = pub_sys_auc_ratio,
    "Systemic % diff" = sys_pct_diff,
    "Liver AUC ratio (simulated)" = liv_auc_ratio,
    "Liver AUC ratio (published)" = pub_liv_auc_ratio,
    "Liver % diff" = liv_pct_diff
  ) %>%
  knitr::kable(digits = 2, caption = "Simulated vs published Table 1 exposure ratios.")
```

```{r table1-assert}
# Deterministic check: zeroRe() removes all random effects, so both sides use
# identical parameters and there is no cohort draw to vary across rxode2
# builds. Exact bounds are correct here.
stopifnot(
  max(abs(cmp$sys_pct_diff)) < 5,
  max(abs(cmp$liv_pct_diff)) < 5
)

# The paper's qualitative conclusion: systemic exposure rises far faster than
# liver exposure across cirrhosis stage.
sev <- cmp[cmp$arm == "Severe", ]
stopifnot(
  sev$sys_auc_ratio > 10,
  sev$liv_auc_ratio < 2,
  sev$sys_auc_ratio / sev$liv_auc_ratio > 5
)
```

Every ratio reproduces within 5% of the published value, and the qualitative
claim (roughly 13-fold systemic vs under 2-fold hepatic) holds.

The mechanism is worth stating explicitly because it is the paper's point.
Reduced hepatic uptake in severe cirrhosis multiplies `t9`/`t10`/`t11` by
`exp(-2.37)`, about a 91% reduction, which is what drives systemic exposure
up 13-fold. Liver *amounts* therefore stay nearly flat, and the reported
liver *concentration* rises only because the functional liver volume shrank
to 61% of normal.

# Concentration-time profiles

```{r fig-profiles, fig.height = 6}
typ %>%
  filter(time <= 72) %>%
  select(arm, time, OCA = Cc, `Glyco-OCA` = Cc_goca, `Tauro-OCA` = Cc_toca) %>%
  pivot_longer(c(OCA, `Glyco-OCA`, `Tauro-OCA`), names_to = "analyte", values_to = "conc") %>%
  ggplot(aes(time, conc, colour = arm)) +
  geom_line() +
  facet_wrap(~analyte, scales = "free_y", ncol = 1) +
  scale_x_continuous(breaks = seq(0, 72, 12)) +
  labs(x = "Time (h)", y = "Plasma concentration (nM)", colour = "Hepatic function",
       title = "Typical-value plasma profiles after a single 10 mg oral dose",
       subtitle = "Compare Edwards 2016 Figures 2 and 3 (visual predictive checks)") +
  theme_bw()
```

The sawtooth on the conjugate profiles is meal-driven gallbladder emptying:
between meals the gallbladder fills through `f22` and cannot discharge,
then releases through `f23` during each 90-minute contraction window.

```{r fig-mass-distribution}
mass_cols <- c("systemic", "portal", "sinusoidal", "liver", "bileduct", "gallbladder", "gut")
mass <- typ %>%
  filter(arm %in% c("Normal", "Severe")) %>%
  mutate(
    systemic    = systemic_oca + systemic_goca + systemic_toca,
    portal      = portal_oca + portal_goca + portal_toca,
    sinusoidal  = sinusoidal_oca + sinusoidal_goca + sinusoidal_toca,
    liver       = liver_oca + liver_goca + liver_toca,
    bileduct    = bileduct_goca + bileduct_toca,
    gallbladder = gallbladder_goca + gallbladder_toca,
    gut         = gut_oca + gut_goca + gut_toca
  ) %>%
  group_by(arm) %>%
  summarise(across(all_of(mass_cols), ~ trapz(time, .x)), .groups = "drop") %>%
  pivot_longer(-arm, names_to = "space", values_to = "auc") %>%
  group_by(arm) %>%
  mutate(pct = 100 * auc / sum(auc)) %>%
  ungroup()

pub_mass <- tibble::tribble(
  ~space,        ~Normal_pub, ~Severe_pub,
  "systemic",    1.03,        10.36,
  "portal",      0.77,        2.33,
  "sinusoidal",  0.05,        0.45,
  "liver",       7.88,        10.44,
  "bileduct",    0.14,        0.13,
  "gallbladder", 39.94,       37.74,
  "gut",         50.20,       38.56
)

wide <- mass %>%
  select(arm, space, pct) %>%
  pivot_wider(names_from = arm, values_from = pct) %>%
  left_join(pub_mass, by = "space")

wide %>%
  select(space, Normal, Normal_pub, Severe, Severe_pub) %>%
  rename("Space" = space, "Normal (simulated)" = Normal, "Normal (published)" = Normal_pub,
         "Severe (simulated)" = Severe, "Severe (published)" = Severe_pub) %>%
  knitr::kable(digits = 2, caption = "Simulated vs published distribution of total OCA across spaces (% of exposure-weighted nanomoles); published values are the Table 1 'Mean Simulated OCA Distribution' block.")
```

Six of the seven spaces track the published distribution closely. Two do not,
for different reasons:

* The **gallbladder / gut split** depends on meal timing, which the paper does
  not publish (see Assumptions and deviations). Their *sum* is 91.4%
  simulated vs 90.1% published, so the pool is right and only its internal
  split moves.
* The **liver** row moves in the opposite direction to the published value
  (falling rather than rising with cirrhosis stage). This reflects an
  inconsistency inside Table 1 itself, quantified in Assumptions and
  deviations below; it is not resolvable from the on-disk sources, and
  nothing was tuned to chase it.

```{r mass-assert}
gut_gb_normal <- sum(wide$Normal[wide$space %in% c("gut", "gallbladder")])
gut_gb_pub <- sum(pub_mass$Normal_pub[pub_mass$space %in% c("gut", "gallbladder")])

# Structural checks only, on the rows that are internally consistent in the
# source. A mis-transcribed volume, flow or dose breaks these immediately.
stopifnot(
  # The enterohepatic pool dominates in health, and its size (not its
  # internal split) is meal-timing independent.
  abs(gut_gb_normal - gut_gb_pub) < 10,
  # Systemic, portal and sinusoidal shares all rise several-fold in
  # Child-Pugh C, as published.
  wide$Severe[wide$space == "systemic"] > 5 * wide$Normal[wide$space == "systemic"],
  wide$Severe[wide$space == "portal"] > 2 * wide$Normal[wide$space == "portal"],
  # And those three rows match the published shares to within 2 percentage
  # points, which is the real transcription check.
  max(abs(wide$Severe[wide$space %in% c("systemic", "portal", "sinusoidal", "bileduct")] -
            wide$Severe_pub[wide$space %in% c("systemic", "portal", "sinusoidal", "bileduct")])) < 2
)
```

# Stochastic simulation and PKNCA validation

The published BSV (195% CV on `Ka`, 78.1% CV on `f22`) is carried here. The
resulting cohort feeds a standard PKNCA analysis of unconjugated OCA.

```{r simulate-stochastic}
rxSetSeed(20260902)
set.seed(20260902)

sim_arm <- function(row) {
  ev <- build_events(row$HEPIMP_MILD, row$HEPIMP_MOD, row$HEPIMP_SEV, N_PER_ARM)
  s <- rxSolve(mod, ev, returnType = "data.frame")
  s$arm <- row$arm
  s
}
pop <- do.call(rbind, lapply(split(arms, seq_len(nrow(arms))), sim_arm))
pop <- pop %>%
  mutate(
    arm = factor(arm, levels = arms$arm),
    uid = paste(arm, id, sep = "-"),
    total_sys_ng = (Cc * MW[["oca"]] + Cc_goca * MW[["goca"]] + Cc_toca * MW[["toca"]]) / 1000
  )
```

```{r pknca}
conc_df <- pop %>%
  select(uid, arm, time, Cc) %>%
  filter(!is.na(Cc))

# Defensive time-zero row: PKNCA warns if the AUC interval starts before the
# first measurement.
conc_df <- bind_rows(
  conc_df,
  conc_df %>% group_by(uid) %>% slice(1) %>% mutate(time = 0, Cc = 0) %>% ungroup()
) %>%
  distinct(uid, time, .keep_all = TRUE) %>%
  arrange(uid, time)

dose_df <- conc_df %>%
  group_by(uid, arm) %>%
  summarise(time = 0, dose = dose_nmol, .groups = "drop")

# PKNCAdose rejects a slash (nested) grouping, so both sides use the crossed
# form. Each uid belongs to exactly one arm, so the partition is identical.
o_conc <- PKNCAconc(conc_df, Cc ~ time | arm + uid)
o_dose <- PKNCAdose(as.data.frame(dose_df), dose ~ time | arm + uid)
o_data <- PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(
    start = 0, end = 216,
    cmax = TRUE, tmax = TRUE, auclast = TRUE, half.life = TRUE
  )
)
res <- suppressWarnings(pk.nca(o_data))
nca <- as.data.frame(res)
```

```{r nca-summary}
nca_summary <- nca %>%
  filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "half.life")) %>%
  group_by(arm, PPTESTCD) %>%
  summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = PPTESTCD, values_from = median)

nca_summary %>%
  rename(
    "Arm" = arm,
    "Cmax (nM)" = cmax,
    "Tmax (h)" = tmax,
    "AUClast (nM*h)" = auclast,
    "t1/2 (h)" = half.life
  ) %>%
  knitr::kable(digits = 2, caption = "PKNCA summary for unconjugated OCA, by hepatic-impairment arm (median across subjects).")
```

```{r nca-assert}
# Robust, centre-based checks. The cohort draw is not reproducible across
# rxode2 versions or thread counts, so these bound the MEDIAN and use
# generous headroom rather than per-subject extremes.
med_auc <- setNames(nca_summary$auclast, as.character(nca_summary$arm))
stopifnot(
  # Unconjugated-OCA exposure must increase monotonically with cirrhosis stage.
  med_auc[["Normal"]] < med_auc[["Mild"]],
  med_auc[["Mild"]] < med_auc[["Moderate"]],
  med_auc[["Moderate"]] < med_auc[["Severe"]],
  # And the severe-vs-normal increase must be large, as the paper reports.
  med_auc[["Severe"]] / med_auc[["Normal"]] > 5,
  all(nca_summary$tmax >= 0)
)
```

# Comparison against the published exposure metrics

Table 1 reports absolute total-OCA exposure alongside the ratios. The
comparison below uses the typical-value solve, so the Cmax rows are expected
to sit low: the published values are means over a 200-replicate population
simulation, and with 195% CV on `Ka` the mean of the per-subject maxima is
substantially above the maximum of the typical-value profile. AUC, which is
far less sensitive to that dispersion, is the informative row.

```{r published-comparison}
pub_abs <- tibble::tribble(
  ~arm,       ~metric,               ~published,
  "Normal",   "Systemic AUC (ng*h/mL)", 2339,
  "Mild",     "Systemic AUC (ng*h/mL)", 3156,
  "Moderate", "Systemic AUC (ng*h/mL)", 18785,
  "Severe",   "Systemic AUC (ng*h/mL)", 30986,
  "Normal",   "Liver AUC (ng*h/mL)",    47427,
  "Mild",     "Liver AUC (ng*h/mL)",    53032,
  "Moderate", "Liver AUC (ng*h/mL)",    69540,
  "Severe",   "Liver AUC (ng*h/mL)",    82521
)

sim_abs <- exposure %>%
  select(arm, `Systemic AUC (ng*h/mL)` = sys_auc, `Liver AUC (ng*h/mL)` = liv_auc) %>%
  pivot_longer(-arm, names_to = "metric", values_to = "simulated")

abs_cmp <- pub_abs %>%
  left_join(sim_abs, by = c("arm", "metric")) %>%
  mutate(
    pct_diff = 100 * (simulated - published) / published,
    flag = ifelse(abs(pct_diff) > 20, "*", "")
  )

abs_cmp %>%
  rename("Arm" = arm, "NCA parameter" = metric, "Published" = published,
         "Simulated" = simulated, "% difference" = pct_diff, " " = flag) %>%
  knitr::kable(digits = 1, caption = "Simulated (typical value) vs published absolute exposure. Rows differing by more than 20% are starred.")
```

```{r abs-assert}
# Deterministic solve again, so a fixed bound is appropriate. Absolute
# exposure is the check that catches a mis-transcribed volume, flow or dose.
stopifnot(max(abs(abs_cmp$pct_diff)) < 25)
```

Both AUC families land within about 12% of the published values across all
four arms, from a model whose parameters were transcribed without any
tuning.

# Assumptions and deviations

* **Meal times are assumed.** The paper states that subjects "received
  standardized meals at specified times during inpatient observation" and
  that gallbladder contraction lasts 90 min from the start of a meal, but it
  does not publish the meal clock times. This vignette assumes two meals per
  day at 4 h and 10 h after the morning dose, following the sibling bile-acid
  model `Zuo_2016_UDCA`. Meal timing shifts the phase of the conjugate
  sawtooth; it does not materially move the AUC comparisons above.
* **`f1`, `f2` and `f5` are derived, not printed.** Figure 1 labels all three
  arrows but Supplemental Table 3 tabulates neither `f1` nor `f5`. Their
  values follow from blood-flow continuity on the portal and sinusoidal
  spaces (`f1 = f2 + f3`, `f5 = f3 + f4`) given the printed `f3` and `f4`;
  `f2` follows from the paper's printed equal-magnitude shunting rule and is
  zero in healthy subjects, consistent with its dotted rendering in Figure 1.
  These are consequences of the paper's own topology and its own printed
  values, not substitutions from an external source.
* **Compartment volumes come from the Figure 1 panel.** The Methods delegate
  volumes to the base CDCA model of Molino 1986 (reference 18), which is
  paywalled and not on disk. The Volumes legend printed inside the Figure 1
  panel reproduces all seven values, so no upstream lookup was needed. An
  earlier extraction attempt deferred this paper on the belief that the
  volumes were unrecoverable; reading the figure panel falsified that.
* **Residual error is the healthy-volunteer set.** The paper estimated a
  second residual-error set on the hepatic-impairment cohort (Supplemental
  Table 4: proportional 122 / 112 / 123%, additive 0.993 / 0.273 / 0.532 nM).
  A single `ini()` block cannot carry two residual-error structures, so the
  healthy-volunteer values from Supplemental Table 3 are encoded, matching
  the larger dataset (8,248 vs 928 samples). The BSV values likewise use the
  healthy-volunteer estimates; the impairment cohort re-estimated 246% and
  168% CV.
* **Parameter count.** The paper states 22 structural parameters ("seven flow
  parameters, four biotransformation parameters, and 11 transport
  parameters") while Supplemental Table 3 tabulates 21 rows. The four
  biotransformation (`b15`, `b16`, `b36`, `b37`) and 11 transport
  (`t9`-`t13`, `t19`, `t21`, `t33`-`t35`, `Ka`) parameters reconcile exactly;
  the flow group is where the counting is ambiguous, because Figure 1 shows
  eight `f` arrows plus `Kout` and Supplemental Table 3 tabulates six of
  them. Every flux in the diagram is nonetheless either tabulated or fixed by
  continuity, so the encoded model is complete.
* **No unconjugated-OCA biliary secretion.** Figure 1 has no liver-to-bile-duct
  arrow in the OCA column, and Supplemental Table 3 has no `t20`. OCA is
  therefore modelled as entering bile only after conjugation.
* **Liver volume does not enter the ODEs.** Every hepatic flux is a
  first-order rate constant acting on amount, so the functional-liver-volume
  mechanism affects only the reported liver concentration. This is what makes
  the paper's headline result arithmetically possible.
* **Table 1 is internally inconsistent on the liver row.** The systemic rows
  fix the ratio of total simulated OCA amount in Child-Pugh C to normal:
  systemic AUC rises 13.2-fold while the systemic *share* rises
  10.36/1.03 = 10.06-fold, so total body amount rises 13.2/10.06 = 1.31-fold.
  Applying that to the liver share gives a liver amount ratio of
  (10.44/7.88) x 1.31 = **1.739**. But Table 1's liver AUC ratio is
  **1.74** - identical to three significant figures - which means the liver
  AUC column behaves as an *amount* ratio, i.e. it was computed without
  applying the functional-liver-volume reduction that Supplemental Table 2
  lists as one of the four impairment mechanisms. Read the other way, with
  the volume reduction applied, the liver AUC ratio implies an amount ratio
  of 1.74 x 0.61 = 1.06, which contradicts the mass-distribution row by a
  factor of 1.64. The two Table 1 blocks cannot both be right.

  The packaged model applies the volume reduction to the reported liver
  concentration (the only place it can act, since every hepatic flux is a
  first-order rate constant on amount). That reproduces the published liver
  AUC ratios 1.12 / 1.47 / 1.74 to within 2%, at the cost of the liver
  mass-share row. No parameter was adjusted to trade one against the other;
  which of the two Table 1 blocks is authoritative cannot be settled from the
  on-disk sources.
* **Supplemental Table 4 notation.** The severe tauro-conjugation row is
  printed as `tb16*`; the intended notation is `b16*`, consistent with the
  mild and moderate rows and with Supplemental Table 2.
* **Reference-list DOI error.** Edwards 2016 reference 18 cites Molino 1986
  as doi:10.1111/j.1365-2362.1986.tb01013.x. The correct DOI is
  doi:10.1111/j.1365-2362.1986.tb01015.x (PMID 3100308); the cited DOI
  resolves to an unrelated 1986 paper on partial hepatectomy in diabetic
  rats.

# Session info

```{r session-info}
sessionInfo()
```
