---
title: "Bictegravir (Ekobena 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bictegravir (Ekobena 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Ekobena_2025_bictegravir"))$reference`
* Description: `r rxode2::rxode(readModelDb("Ekobena_2025_bictegravir"))$description`
* Article: <https://doi.org/10.1093/jac/dkaf297>
* Open-access full text: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC12494133/>

Bictegravir is an integrase strand transfer inhibitor given only as part of the
fixed-dose bictegravir / emtricitabine / tenofovir alafenamide combination. This
is the first description of its population PK in a real-world, unselected
outpatient setting; every previous analysis was the manufacturer's, which is not
published in the scientific literature.

# Population

The model was built on `r readModelDb("Ekobena_2025_bictegravir")()$population$n_observations`
steady-state plasma concentrations from
`r readModelDb("Ekobena_2025_bictegravir")()$population$n_subjects` people with
HIV enrolled in the Swiss HIV Cohort Study and followed by therapeutic drug
monitoring at Lausanne University Hospital between July 2019 and July 2024
(Ekobena 2025 Table 1). Median age was 51 years (range 19-81), median body
weight 74 kg (range 37-135), median BMI 24.5 kg/m^2 (12.5-44.5) and 31% of
participants were female. Reported ethnicity was 65% White, 26% Black, 4%
Hispanic American, 4% Asian and 1% other or missing.

This is a sparse therapeutic-drug-monitoring dataset, not a rich PK study, and
that shapes what the model can and cannot do. Of 815 concentrations collected,
107 were excluded (42 undetectable, most likely non-adherence; 40 missing time
information; 25 with uncertainties about the reported information). A median of
**one** sample was taken per individual (range 1-4), at 0.25-32 h after the last
dose. Steady state was assumed for every sample on the strength of the long
follow-up and a reference half-life of 17.3 h.

The same information is available programmatically via the model's `population`
metadata:

```{r population}
str(readModelDb("Ekobena_2025_bictegravir")()$population, max.level = 1)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Ekobena_2025_bictegravir.R`. The table
below collects them in one place for review. Every value comes from the "Final
model Estimate (RSE, %)" column of Ekobena 2025 Table 2; the bootstrap medians
in the adjacent column agree with all of them to within 3%.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` (fixed)  | `log(0.64)` 1/h | Table 2, row `ka (h-1)` = "0.64 FIX"; Methods explains it was fixed because of sparse early post-dose data (initial RSE 38%) and bootstrap instability |
| `lcl`          | `log(0.46)` L/h | Table 2, row `CL (L/h)` = 0.46 (RSE 1%); bootstrap 0.46 [0.44-0.47] |
| `lvc`          | `log(10.9)` L | Table 2, row `V (L)` = 10.9 (RSE 5%); bootstrap 11.2 [10.1-12.4] |
| `e_wt_cl`      | 0.35 | Table 2, row `beta Bodyweight` = 0.35 (RSE 19%); bootstrap 0.36 [0.22-0.49] |
| `e_age_cl`     | -0.20 | Table 2, row `beta Age` = -0.20 (RSE 28%); bootstrap -0.19 [-0.31 to -0.08] |
| `etalcl`       | 0.04727 (variance) | Table 2, row `omega CL (CV%)` = 22 (RSE 6%); `log(1 + 0.22^2)` |
| `propSd`       | 0.27 | Table 2, row `sigma prop (CV%)` = 27 (RSE 4%) |
| Reference body weight `BWRef` | 70 kg | Results, sentence immediately below the final covariate equation |
| Median age `AgeM` | 51 years | Results, same sentence; also Table 1 median age |
| `d/dt(depot)`, `d/dt(central)` | n/a | Results, "A one-compartment model with linear elimination, parameterized in terms CL, V, ka fixed to 0.64 h-1"; a second compartment was rejected (dBICc = +18.6) |
| `cl` covariate equation | n/a | Results, the displayed equation `CLi = CL * exp(beta_Bodyweight * log(BW/BWRef) + beta_Age * (Age - AgeM)/AgeM + eta_i)` |
| IIV on CL only, none on V | n/a | Results, "Assignment of IIV on V did not improve data description (dBICc = +8.4)" |
| Proportional residual error | n/a | Results, "The RUV was best captured by a proportional error model" |

# Structural checks against closed-form solutions

Before simulating a cohort, verify that the packaged model reproduces the
analytic one-compartment first-order-absorption solution and the derived
quantities the paper quotes. These checks compare two sides that use the *same*
parameter values, so the only difference is numerical integration error and the
bounds are correspondingly tight.

```{r structural}
mod <- readModelDb("Ekobena_2025_bictegravir")
mod_typ <- rxode2::zeroRe(mod)

# Parameters at the covariate reference point (70 kg, 51 years).
CL0 <- 0.46; V <- 10.9; KA <- 0.64
kel <- CL0 / V
tau <- 24                      # 50 mg once daily
n_doses <- 20L                 # ~29 half-lives; comfortably steady state
t_last <- tau * (n_doses - 1L)

# Analytic steady-state concentration for a one-compartment model with
# first-order absorption (dose in mg, V in L -> mg/L; x1000 -> ng/mL).
css <- function(t, cl, v = V, ka = KA, dose = 50, ii = tau) {
  k <- cl / v
  dose * ka / (v * (ka - k)) *
    (exp(-k * t) / (1 - exp(-k * ii)) - exp(-ka * t) / (1 - exp(-ka * ii))) * 1000
}

structural <- tibble::tibble(
  quantity = c("Elimination rate constant kel (1/h)",
               "Terminal half-life (h)",
               "Single-dose Tmax (h)",
               "Steady-state Tmax (h)",
               "Steady-state AUC0-tau (ng*h/mL)"),
  value = c(kel,
            log(2) / kel,
            log(KA / kel) / (KA - kel),
            log((KA / kel) * (1 - exp(-kel * tau)) / (1 - exp(-KA * tau))) / (KA - kel),
            50 / CL0 * 1000)
)
knitr::kable(structural, digits = 4,
             caption = "Derived quantities at the covariate reference point.")
```

The paper's base model reports a Tmax of 4.5 h and a terminal half-life of
16.3 h (Results, "Base and covariate model"). The final-model parameters give
`r sprintf("%.2f", log(KA / kel) / (KA - kel))` h and
`r sprintf("%.1f", log(2) / kel)` h, so the packaged parameters are consistent
with the quantities the authors quote. Note that the paper's 4.5 h is the
*single-dose* Tmax; at steady state the accumulated trough shifts the peak
earlier, to about `r sprintf("%.1f", log((KA / kel) * (1 - exp(-kel * tau)) / (1 - exp(-KA * tau))) / (KA - kel))` h.

```{r structural-gate}
# Solve the packaged ODE model at typical values and compare against the
# closed form. Same parameters on both sides, so this is pure integration
# error -- a tight bound is correct here.
ev_typ <- rxode2::et(amt = 50, time = tau * (seq_len(n_doses) - 1L), cmt = "depot") |>
  rxode2::et(t_last + seq(0, tau, by = 0.25), cmt = "central")

sim_typ <- rxode2::rxSolve(mod_typ, ev_typ, params = c(WT = 70, AGE = 51),
                           returnType = "data.frame")
sim_typ <- dplyr::mutate(sim_typ, t_rel = time - t_last, Cc_ng = Cc * 1000)

chk <- sim_typ |>
  dplyr::filter(!is.na(Cc_ng)) |>
  dplyr::mutate(analytic = css(t_rel, CL0),
                pct_diff = 100 * (Cc_ng - analytic) / analytic)

stopifnot(
  # Numerical-integration agreement over the whole dosing interval.
  max(abs(chk$pct_diff)) < 0.05,
  # The full 50 mg dose enters the depot: the model has no bioavailability
  # term, because CL and V are already apparent (CL/F, V/F) values. At steady
  # state the depot also carries the residue of the previous dose, so the
  # analytic peak is 50 / (1 - exp(-ka * tau)), not exactly 50.
  isTRUE(all.equal(max(sim_typ$depot), 50 / (1 - exp(-KA * tau)),
                   tolerance = 1e-6))
)
sprintf("Max |%% difference| vs closed form over the dosing interval: %.4f%%",
        max(abs(chk$pct_diff)))
```

## Covariate-effect regression test

The Results section states two exact numbers that pin the functional form of
both covariate effects. These are algebraic identities of the packaged
coefficients, so they are gated tightly.

```{r covariate-gate}
cl_i <- function(wt, age) {
  CL0 * exp(0.35 * log(wt / 70) - 0.20 * (age - 51) / 51)
}

cov_chk <- tibble::tibble(
  claim = c("BW 100 kg, age 51 y: CL increased by 13%",
            "BW 70 kg, age 80 y: CL decreased by 11%"),
  published_pct = c(13, -11),
  model_pct = c(100 * (cl_i(100, 51) / CL0 - 1),
                100 * (cl_i(70, 80) / CL0 - 1))
)
knitr::kable(cov_chk, digits = 2,
             caption = "Ekobena 2025 Results: covariate effects on apparent clearance.")

stopifnot(
  # The paper rounds to whole percent, so 0.5 pp is the exact tolerance.
  all(abs(cov_chk$model_pct - cov_chk$published_pct) < 0.5)
)
```

This is the check that catches the single easiest transcription error in this
model: the age term is linear in the **fractional** deviation
`(AGE - 51) / 51`, not in the raw deviation in years. Dropping the division by
51 would turn the second row into a 99.7% decrease.

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the six scenarios the paper simulates in its Table 3: continuous covariates
drawn from a uniform distribution over each clinically meaningful band, plus
between-subject variability on apparent clearance.

Two deliberate departures from the paper's simulation setup:

* **200 subjects per arm rather than 1000.** This is the nlmixr2lib cap; it is
  ample for a median and keeps the vignette inside the render time budget.
* **A deterministic lattice instead of random draws.** Each arm's 200 subjects
  take between-subject `eta` values on a quantile lattice
  `qnorm((j - 0.5) / 200) * omega`, and covariates on uniform lattices permuted
  by a fixed base-R seed. Nothing here touches rxode2's simulation RNG, whose
  streams are partitioned per solver thread and therefore differ between a
  2-core CI runner and a 16-thread workstation. The cohort below is
  bit-identical everywhere, so the assertions downstream are reproducible rather
  than merely lucky. The trade-off is that a 200-point lattice truncates the
  extreme tails, so the minimum and maximum columns of the paper's Table 3 are
  reported but not gated.

```{r cohort}
n_arm <- 200L
omega <- sqrt(log(1 + 0.22^2))   # Table 2 reports omega_CL as 22 CV%
lcl0 <- log(CL0)

RNGkind(sample.kind = "Rejection")   # pin sample() across R versions
set.seed(20250909)
perm_age <- sample.int(n_arm)
perm_wt <- sample.int(n_arm)

lattice <- function(perm, lo, hi) lo + (perm - 0.5) / n_arm * (hi - lo)

arms <- tibble::tribble(
  ~group,                       ~age_lo, ~age_hi, ~wt_lo, ~wt_hi,
  "Typical patient",                 51,      51,     70,     70,
  "Age 20-65 y; BW 95-135 kg",       20,      65,     95,    135,
  "Age 65-80 y; BW 95-135 kg",       65,      80,     95,    135,
  "Age 20-65 y; BW 40-95 kg",        20,      65,     40,     95,
  "Age 65-80 y; BW 40-95 kg",        65,      80,     40,     95,
  "Age 20 y; BW 135 kg",             20,      20,    135,    135
)

make_arm <- function(k) {
  a <- arms[k, ]
  tibble::tibble(
    id = (k - 1L) * n_arm + seq_len(n_arm),
    group = a$group,
    AGE = lattice(perm_age, a$age_lo, a$age_hi),
    WT = lattice(perm_wt, a$wt_lo, a$wt_hi),
    # The between-subject eta is folded into a per-subject `lcl` and solved
    # with zeroRe(), which is algebraically identical to sampling etalcl but
    # fully deterministic.
    lcl = lcl0 + qnorm((seq_len(n_arm) - 0.5) / n_arm) * omega
  )
}
cohort <- dplyr::bind_rows(lapply(seq_len(nrow(arms)), make_arm))

# Dosing + observation template, expanded per subject.
template <- as.data.frame(
  rxode2::et(amt = 50, time = tau * (seq_len(n_doses) - 1L), cmt = "depot") |>
    rxode2::et(t_last + seq(0, tau, by = 0.25), cmt = "central")
)

events <- cohort |>
  dplyr::rowwise() |>
  dplyr::reframe(dplyr::bind_cols(id = id, group = group, AGE = AGE, WT = WT,
                                  lcl = lcl, template)) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

# Disjoint IDs across arms; duplicate IDs would silently merge subjects.
stopifnot(
  nrow(cohort) == n_arm * nrow(arms),
  !anyDuplicated(cohort$id),
  !anyDuplicated(unique(events[, c("id", "time", "evid")]))
)
```

# Simulation

```{r simulate}
sim <- as.data.frame(
  rxode2::rxSolve(mod_typ, events = events, keep = c("group"))
) |>
  dplyr::mutate(t_rel = time - t_last, Cc = Cc * 1000)   # mg/L -> ng/mL

# rxSolve returns observation records only -- there is no `evid` column on the
# output, so no filtering is needed here.
stopifnot(
  # A large residual error can drive simulated concentrations negative and
  # silently turn AUC into NaN; there is no residual error on Cc, but assert
  # positivity and completeness anyway before any NCA runs on this frame.
  !anyNA(sim$Cc),
  all(sim$Cc > 0),
  nrow(sim) == nrow(cohort) * length(t_last + seq(0, tau, by = 0.25))
)
```

# Replicate published figures

```{r figure-1a, fig.width = 7, fig.height = 4.5}
# Replicates Figure 1(a) of Ekobena 2025: simulated steady-state percentiles
# for the typical patient (51 years old, 70 kg) under 50 mg once daily.
sim |>
  dplyr::filter(group == "Typical patient") |>
  dplyr::group_by(t_rel) |>
  dplyr::summarise(Q05 = quantile(Cc, 0.05), Q25 = quantile(Cc, 0.25),
                   Q50 = median(Cc), Q75 = quantile(Cc, 0.75),
                   Q95 = quantile(Cc, 0.95), .groups = "drop") |>
  ggplot(aes(t_rel, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.18) +
  geom_ribbon(aes(ymin = Q25, ymax = Q75), alpha = 0.30) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 760, linetype = "dotted") +
  labs(x = "Time after dose (h)", y = "Bictegravir concentration (ng/mL)",
       title = "Figure 1(a) - typical patient at steady state",
       caption = paste("Replicates Figure 1(a) of Ekobena 2025.",
                       "Dotted line: 760 ng/mL minimum concentration",
                       "recommended for efficacy."))
```

```{r figure-1b, fig.width = 7, fig.height = 4.5}
# Replicates Figure 1(b) of Ekobena 2025: simulated median concentrations for
# each covariate group.
sim |>
  dplyr::group_by(group, t_rel) |>
  dplyr::summarise(Q50 = median(Cc), .groups = "drop") |>
  ggplot(aes(t_rel, Q50, colour = group)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 760, linetype = "dotted") +
  labs(x = "Time after dose (h)", y = "Median concentration (ng/mL)",
       colour = NULL,
       title = "Figure 1(b) - median profile by covariate group",
       caption = "Replicates Figure 1(b) of Ekobena 2025.") +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 3))
```

# PKNCA validation

NCA is run over the final steady-state dosing interval. For a once-daily
extravascular profile at steady state, `cmin` over `[t_last, t_last + tau]` is
the trough concentration the paper tabulates: the interval starts at a dose, so
its two endpoints are both troughs and are equal at steady state. PKNCA's
dedicated `ctrough` parameter is not used here because it returns `NA` for this
interval definition.

```{r pknca}
# Only `!is.na(Cc)` in the filter -- adding `time > 0` or `Cc > 0` would drop
# the anchor record PKNCA needs.
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, group)

dose_df <- events |>
  dplyr::filter(evid == 1, time == t_last) |>
  dplyr::select(id, time, amt, group)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | group + id,
                             concu = "ng/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | group + id, doseu = "mg")

intervals <- data.frame(
  start = t_last, end = t_last + tau,
  cmax = TRUE, tmax = TRUE, cmin = TRUE,
  auclast = TRUE, cav = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
```

```{r pknca-massbalance}
# Mass-balance gate on typical values: at steady state AUC0-tau must equal
# Dose / CL exactly. This is the cheapest guard against a mis-parameterised
# clearance or an accidentally auto-solved linCmt() model.
auc_typ <- as.data.frame(nca_res$result) |>
  dplyr::filter(group == "Typical patient", PPTESTCD == "auclast")

stopifnot(
  # Median over the arm sits at eta = 0 exactly, so this compares against the
  # typical-value analytic result; trapezoidal error on a 0.25 h grid only.
  abs(median(auc_typ$PPORRES) / (50 / CL0 * 1000) - 1) < 0.005,
  # linCmt() would have discarded the explicit ODEs; assert they survived.
  identical(rxode2::rxode(mod)$state, c("depot", "central"))
)
sprintf("Median steady-state AUC0-tau: %.0f ng*h/mL; Dose/CL = %.0f ng*h/mL",
        median(auc_typ$PPORRES), 50 / CL0 * 1000)
```

## Comparison against published NCA

Ekobena 2025 Table 3 reports simulated steady-state trough concentrations under
the standard 50 mg once-daily regimen for six covariate scenarios. It is the
paper's only tabulated exposure output, and reproducing all six rows is a
six-point regression test on the structural parameters *and* both covariate
coefficients simultaneously.

```{r nca_compare}
published <- tibble::tribble(
  ~group,                       ~cmin,
  "Typical patient",             2781,
  "Age 20-65 y; BW 95-135 kg",   1991,
  "Age 65-80 y; BW 95-135 kg",   2436,
  "Age 20-65 y; BW 40-95 kg",    2760,
  "Age 65-80 y; BW 40-95 kg",    3267,
  "Age 20 y; BW 135 kg",         1593
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "group",
  params = "cmin",
  units = c(cmin = "ng/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste("Simulated vs. published steady-state trough concentration",
                  "(Ekobena 2025 Table 3). * differs from reference by >20%."),
  digits = 1
)
```

```{r nca_compare_gate}
sim_cmin <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD == "cmin") |>
  dplyr::group_by(group) |>
  dplyr::summarise(sim = median(PPORRES), .groups = "drop") |>
  dplyr::left_join(published, by = "group") |>
  dplyr::mutate(pct = 100 * (sim / cmin - 1))

knitr::kable(
  sim_cmin |>
    dplyr::rename("Group" = group, "Simulated Ctrough (ng/mL)" = sim,
                  "Published Ctrough (ng/mL)" = cmin, "Difference (%)" = pct),
  digits = 1,
  caption = "Per-group deviation from Ekobena 2025 Table 3."
)

# The two fixed-covariate arms ("Typical patient", "Age 20 y; BW 135 kg") have
# no covariate spread, so the lattice median lands exactly on the typical value
# and the only residual is Monte Carlo noise in the PAPER's own 1000-subject
# simulation. The four banded arms carry an additional, irreducible difference:
# the median of a nonlinear function over a uniform covariate band depends on
# how the covariate draws pair with the clearance draws, and the paper's random
# pairing is not recoverable. Observed deviations at authoring were 0.7% and
# 1.3% (fixed arms) and 1.0-5.0% (banded arms).
fixed_arms <- c("Typical patient", "Age 20 y; BW 135 kg")
stopifnot(
  all(abs(sim_cmin$pct[sim_cmin$group %in% fixed_arms]) < 5),
  all(abs(sim_cmin$pct) < 12),
  # A gate that cannot go red is worthless: substituting an allometric 0.75 for
  # the estimated 0.35 weight exponent moves the heaviest arm by ~25%, and
  # dropping the /51 on the age term moves it by orders of magnitude.
  nrow(sim_cmin) == 6L, !anyNA(sim_cmin$pct)
)
```

## Derived claims in the Results and Discussion

The paper makes four further quantitative statements about its simulations.
These are summary claims over the same six arms, quoted to the nearest whole
percent, so they are reported here and gated only on the directional content
that the model must reproduce.

```{r derived-claims}
med <- setNames(sim_cmin$sim, sim_cmin$group)

claims <- tibble::tibble(
  claim = c(
    "Age 65-80 vs 20-65, BW 95-135 kg: higher trough in older",
    "Age 65-80 vs 20-65, BW 40-95 kg: higher trough in older",
    "BW 95-135 vs 40-95 kg (age 20-65): lower trough in heavier",
    "20-year-old, 135 kg vs typical patient: lower trough"
  ),
  published_pct = c(20, 20, -27, -43),
  model_pct = c(
    100 * (med[["Age 65-80 y; BW 95-135 kg"]] / med[["Age 20-65 y; BW 95-135 kg"]] - 1),
    100 * (med[["Age 65-80 y; BW 40-95 kg"]] / med[["Age 20-65 y; BW 40-95 kg"]] - 1),
    100 * (med[["Age 20-65 y; BW 95-135 kg"]] / med[["Age 20-65 y; BW 40-95 kg"]] - 1),
    100 * (med[["Age 20 y; BW 135 kg"]] / med[["Typical patient"]] - 1)
  )
)
knitr::kable(claims, digits = 1,
             caption = "Simulation claims in Ekobena 2025 Results and Discussion.")

stopifnot(
  # Direction is structural and must hold: age lowers clearance (raises
  # exposure), weight raises clearance (lowers exposure).
  all(claims$model_pct[1:2] > 0),
  all(claims$model_pct[3:4] < 0),
  # Magnitudes within 8 percentage points of the paper's rounded figures.
  all(abs(claims$model_pct - claims$published_pct) < 8)
)
```

```{r subtherapeutic}
# Discussion: "20-year-old individuals weighing 135 kg could have a median 43%
# decrease in bictegravir Ctrough, relative to the typical individual, with
# almost 5% of Ctrough expected to fall below the minimum recommended
# concentration for efficacy (i.e. 760 ng/mL)."
below <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD == "cmin") |>
  dplyr::group_by(group) |>
  dplyr::summarise(pct_below_760 = 100 * mean(PPORRES < 760), .groups = "drop")

knitr::kable(below, digits = 1,
             caption = "Percentage of simulated troughs below the 760 ng/mL efficacy threshold.")

stopifnot(
  # The typical patient must stay above the threshold; the young, heavy arm
  # must not. Both are directional claims the paper makes explicitly.
  below$pct_below_760[below$group == "Typical patient"] == 0,
  below$pct_below_760[below$group == "Age 20 y; BW 135 kg"] > 0
)
```

# Assumptions and deviations

* **Cohort construction.** The paper simulates 1000 individuals per scenario
  with covariates drawn from uniform distributions; this vignette uses 200 per
  arm (the nlmixr2lib cap) on a deterministic quantile lattice. The lattice
  makes every number here reproducible across machines and rxode2 versions, but
  it truncates the extreme tails, so the `Minimum` and `Maximum` columns of the
  paper's Table 3 are not reproduced or gated. The banded arms additionally
  depend on how covariate draws pair with clearance draws, which the paper does
  not report; observed deviations from Table 3 were 1.0-5.0% for those arms.

* **Residual error in the simulations.** The paper does not state whether its
  Table 3 percentiles and Figure 1 prediction intervals include the
  proportional residual error or only between-subject variability. The
  simulations here carry between-subject variability only. Medians are
  unaffected either way; the 5th and 95th percentiles would widen if residual
  error were added. This is also why the "almost 5% below 760 ng/mL" claim is
  gated only as "greater than zero" rather than on its value.

* **`omega_CL` scale.** Table 2 heads the row `omega CL (CV%)` = 22, so 22 is
  read as a coefficient of variation and converted exactly via
  `omega^2 = log(1 + 0.22^2) = 0.04727`. If the authors instead printed `omega`
  itself under that header -- a common Monolix table convention -- the variance
  would be 0.0484. The difference is 1.2% in the standard deviation and moves
  no gate in this vignette. The sigma row cannot arbitrate the question: for a
  proportional residual error the tabulated CV% *is* the standard deviation, so
  it carries no information about the log-normal convention.

* **Absorption is not informed by the data.** `ka` is fixed at 0.64 1/h and
  carries no between-subject variability, because the study had almost no
  samples shortly after dosing (initial RSE 38%, bootstrap unstable). The
  Discussion says as much. Cmax and Tmax predictions from this model should be
  treated as structural consequences of a fixed constant, not as fitted
  quantities; the trough predictions, which is what the paper uses the model
  for, are robust to it. No absorption lag time was estimated, unlike the
  manufacturer's unpublished analysis.

* **No bioavailability term.** `CL` and `V` are apparent (`CL/F`, `V/F`) values.
  The study has no intravenous reference arm and `F` is neither estimated nor
  reported, so no `f(depot)` is applied -- adding one would double-count the
  unknown bioavailability.

* **Drug-drug interactions are out of scope.** Only 15 participants (2%) took a
  metabolism inhibitor and 4 (<1%) an inducer, and no relationship with
  bictegravir concentrations was detected. The authors attribute this to lack of
  power rather than to absence of an interaction, and note that dual
  CYP3A4/UGT1A1 inhibitors are predicted to raise exposure 2.5- to 2.9-fold and
  strong inducers to cut it by up to 70%. This model must not be used to predict
  exposure under those conditions.

* **Covariates screened but not retained.** BMI was significant univariately
  (dBICc = -3.2) but dropped at backward deletion in favour of body weight;
  serum albumin was considered because bictegravir is >99% protein bound but was
  available for only 34% of drug levels and too tightly clustered to test. Both
  are recorded in the model file's `covariatesDataExcluded` list with the
  paper's reasoning. No coefficient is published for either, so neither can be
  encoded.

* **Reference weight is not the cohort median.** `BWRef` is 70 kg, the rounded
  reference the paper names below its covariate equation, while Table 1 gives a
  cohort median body weight of 74 kg. `AgeM`, by contrast, is 51 years and *is*
  the cohort median.

* **Supplementary material.** Figures S1 (goodness-of-fit) and S2 (pcVPC) are
  available as Supplementary data at JAC Online. They are diagnostic plots and
  contain no parameter values, so nothing in the model file depends on them and
  they were not retrieved.

* **No external validation.** The authors state that no external validation of
  the model could be performed, and that the data come from a moderately diverse
  population with a substantial proportion of missing covariate values.
