---
title: "Recurrent ischemic stroke after a first stroke (Elhefnawy 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recurrent ischemic stroke after a first stroke (Elhefnawy 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Elhefnawy ME, Sheikh Ghadzi SM, Albitar O, Tangiisuran B, Zainal H,
  Looi I, Sidek NN, Aziz ZA, Harun SN. Predictive model of recurrent ischemic
  stroke: model development from real-world data. Front Neurol.
  2023;14:1118711. doi:10.3389/fneur.2023.1118711.
* Article: <https://doi.org/10.3389/fneur.2023.1118711> (Open Access, CC BY)
* Model file: `modellib("Elhefnawy_2023_recurrent_ischemic_stroke")`

This is a parametric time-to-event (TTE) model, not a PK or PK-PD model. There
is no drug exposure term: secondary prevention enters only as a binary
"antiplatelet prescribed at discharge" indicator. Time is in years.

**Read the Errata section before using this model.** The published parameters
do not reproduce the paper's own event counts or its own Kaplan-Meier visual
predictive check; the arithmetic is set out there in full. Everything in this
file is transcribed exactly as published and nothing has been tuned.

```{r model}
mod <- modellib("Elhefnawy_2023_recurrent_ischemic_stroke")
mod
```

# Population

National Neurology Registry (NNEUR) of Malaysia, a multicentre hospital-based
registry covering 13 states. All patients with an index ischemic stroke
registered from August 2009 to December 2016 were extracted: 7,697 adults over
18 years of age, of whom 333 (4.32%) had at least one recurrent ischemic stroke
within the maximum 7.37 years of follow-up. Median time to first recurrence was
1.2 years; 108 of the 333 recurrences (31.43%) happened within the first 6
months, and 36 patients went on to a second recurrence. Median age at the index
stroke was 63.47 years and 55.72% of the cohort were women.

Stroke was diagnosed by WHO criteria and confirmed by brain CT or MRI.
Non-Malaysian citizens and non-ischemic diagnoses were excluded.

Baseline prevalences used below are computed from the Elhefnawy 2023 Table 1
counts, which reconcile exactly with the Results text (e.g. hyperlipidemia
159 + 1,869 = 2,028 = 26.34%).

```{r prevalence}
prev <- tibble::tribble(
  ~covariate,              ~recurrent, ~non_recurrent,
  "DIS_HYPERLIP",                 159,           1869,
  "DIS_IHD",                       77,            802,
  "DIS_HYPERT",                   288,           5218,
  "CONMED_ANTIPLATELET",          285,           6613
) |>
  dplyr::mutate(n = recurrent + non_recurrent, prevalence = n / 7697)

prev |>
  dplyr::rename(
    "Covariate" = covariate,
    "Recurrent IS (n = 333)" = recurrent,
    "No recurrent IS (n = 7,364)" = non_recurrent,
    "Total (n = 7,697)" = n,
    "Prevalence" = prevalence
  ) |>
  knitr::kable(digits = 4)
```

# Source trace

Every value in `ini()` and every equation in `model()`, with the place in
Elhefnawy 2023 it came from.

| Model element | Source location | Published value |
|---|---|---|
| `S(t) = exp(-cumhaz)` | Methods, Equation 1 | `S(t) = e^{-int_0^t h(t)dt}` |
| Gompertz hazard form | Methods, Equation 3; Table 2 row 4 | `h(t) = theta_x * e^{(theta_y) t}` |
| Piecewise scale / shape switching | Table 2 footnote | `theta_x = theta1 if t < 0.5 y, theta3 if t >= 0.5`; `theta_y = theta2 if t < 3 y, theta4 if t >= 3` |
| Covariate model | Methods, unnumbered equation after Eq. 5 | `h(t) = h0(t) * exp(b1 X1 + ... + bn Xn)` |
| `lh0_early_ris` | Table 3 row 1 (theta1, `<6 months`) | 0.238 /year, RSE 19.92% |
| `lh0_late_ris` | Table 3 row 2 (theta3, `>=6 months`) | 0.0016 /year, RSE 21.62% |
| `shape_early_ris` | Table 3 row 3 (theta2, alpha `<3`) | 1.63 /year, RSE 4.81% |
| `shape_late_ris` | Table 3 row 4 (theta4, alpha `>=3`) | 0.23 /year, RSE 20.19% |
| `e_dis_hyperlip_ris` | Table 3 row 5 (theta5) | 0.799, aHR 2.22 (1.81-2.72), RSE 12.89% |
| `e_dis_ihd_ris` | Table 3 row 6 (theta6) | 0.745, aHR 2.10 (1.64-2.69), RSE 16.85% |
| `e_dis_hypert_ris` | Table 3 row 7 (theta7) | 0.711, aHR 2.03 (1.52-2.71), RSE 20.62% |
| `e_conmed_antiplatelet_ris` | Table 3 row 8 (theta8) | -0.514, aHR 0.59 (printed "0.79-0.44"), RSE 28.41% |
| `tbreak_scale_ris` | Table 2 footnote; Table 3 row labels | 0.5 year (fixed, not estimated) |
| `tbreak_shape_ris` | Table 2 footnote; Table 3 row labels | 3 years (fixed, not estimated) |
| `etalh0_*_ris` | Methods, "Development of the base model" | BSV stated to be estimated; **magnitude never reported** -- fixed at 0 |
| `addSd` | not in the source | placeholder for the nlmixr2 likelihood; see Assumptions |

# Covariate patterns

The model has four binary covariates, so a cohort is fully described by the 16
covariate patterns and their weights. Enumerating them exactly, rather than
drawing a random cohort, makes every number below deterministic and identical
across rxode2 versions and thread counts.

Elhefnawy 2023 does not publish the joint distribution of the four covariates,
and its Limitations section states that "the comorbidities were analyzed
independently". The pattern weights below therefore assume independence -- see
Assumptions and deviations.

```{r patterns}
p <- setNames(prev$prevalence, prev$covariate)

patterns <-
  tidyr::expand_grid(
    DIS_HYPERLIP        = 0:1,
    DIS_IHD             = 0:1,
    DIS_HYPERT          = 0:1,
    CONMED_ANTIPLATELET = 0:1
  ) |>
  dplyr::mutate(
    weight =
      ifelse(DIS_HYPERLIP        == 1, p[["DIS_HYPERLIP"]],        1 - p[["DIS_HYPERLIP"]]) *
      ifelse(DIS_IHD             == 1, p[["DIS_IHD"]],             1 - p[["DIS_IHD"]]) *
      ifelse(DIS_HYPERT          == 1, p[["DIS_HYPERT"]],          1 - p[["DIS_HYPERT"]]) *
      ifelse(CONMED_ANTIPLATELET == 1, p[["CONMED_ANTIPLATELET"]], 1 - p[["CONMED_ANTIPLATELET"]]),
    id = dplyr::row_number()
  )

stopifnot(nrow(patterns) == 16L, abs(sum(patterns$weight) - 1) < 1e-12)
```

```{r solve}
obs_times <- sort(unique(c(seq(0, 7.37, by = 0.01), 0.5, 3, 7.37)))

events <-
  patterns |>
  dplyr::select(id, DIS_HYPERLIP, DIS_IHD, DIS_HYPERT, CONMED_ANTIPLATELET) |>
  tidyr::expand_grid(time = obs_times) |>
  dplyr::mutate(evid = 0, amt = 0) |>
  dplyr::arrange(id, time)

# omega = NA / sigma = NA: the etas are fixed at zero, and rxSolve would
# otherwise reuse an omega left over from a previous solve in the session.
sim <-
  rxode2::rxSolve(mod, events, omega = NA, sigma = NA, returnType = "data.frame") |>
  dplyr::as_tibble() |>
  dplyr::mutate(id = as.integer(as.character(id))) |>
  dplyr::left_join(dplyr::select(patterns, id, weight), by = "id")
```

# Replicate Figure 1 -- baseline hazard by time interval

Figure 1 of Elhefnawy 2023 plots the baseline hazard (all predictors set to
zero) as a step: 0.238 over the first 6 months, dropping to essentially zero
afterwards, and flat out to 7 years. That figure plots the *scale* parameter
`theta_x` alone. The hazard the model actually integrates carries the Gompertz
factor `exp(theta_y * t)` as well, so it rises within each interval; both are
shown.

```{r fig1, fig.cap = "Replicates Figure 1 of Elhefnawy 2023 (baseline hazard, all predictors zero). The step is the theta_x scale the paper plots; the curve is the full Gompertz hazard the model integrates."}
base <- dplyr::filter(
  sim,
  DIS_HYPERLIP == 0, DIS_IHD == 0, DIS_HYPERT == 0, CONMED_ANTIPLATELET == 0
)

scale_step <- dplyr::mutate(base, scale_only = ifelse(time < 0.5, 0.238, 0.0016))

ggplot(scale_step, aes(x = time)) +
  geom_line(aes(y = scale_only, colour = "theta_x scale only (paper Figure 1)")) +
  geom_line(aes(y = hazard_ris, colour = "full Gompertz hazard")) +
  scale_colour_manual(values = c("theta_x scale only (paper Figure 1)" = "grey40",
                                 "full Gompertz hazard" = "steelblue")) +
  labs(x = "Years after index ischemic stroke", y = "Hazard (1/year)",
       colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

The step heights are the transcribed Table 3 values exactly, and the two
switch points behave as the Table 2 footnote specifies (`theta1` applies while
`t < 0.5`, `theta3` from `t >= 0.5`).

```{r fig1-checks}
h_at <- function(tt) base$hazard_ris[which.min(abs(base$time - tt))]

stopifnot(
  # t = 0: hazard is exactly theta1, because exp(theta2 * 0) = 1.
  abs(h_at(0) - 0.238) < 1e-9,
  # Each of the three regimes must equal its own published formula exactly.
  # Scale switch at 0.5 y: theta1 applies at 0.49, theta3 from 0.50.
  abs(h_at(0.49) - 0.238  * exp(1.63 * 0.49)) < 1e-9,
  abs(h_at(0.50) - 0.0016 * exp(1.63 * 0.50)) < 1e-9,
  # Shape switch at 3 y: theta2 applies at 2.99, theta4 from 3.00.
  abs(h_at(2.99) - 0.0016 * exp(1.63 * 2.99)) < 1e-9,
  abs(h_at(3.00) - 0.0016 * exp(0.23 * 3.00)) < 1e-9
)
```

# Replicate Figure 2 -- adjusted hazard ratios

Figure 2 and Table 3 report the adjusted hazard ratio for each retained
covariate. Because all four covariates are 0/1 indicators, the hazard ratio is
exactly `exp(coefficient)`, so this is an exact identity check on the
transcription rather than a simulation.

```{r fig2}
coefs <- c(
  DIS_HYPERLIP        = 0.799,
  DIS_IHD             = 0.745,
  DIS_HYPERT          = 0.711,
  CONMED_ANTIPLATELET = -0.514
)
published_hr <- c(
  DIS_HYPERLIP = 2.22, DIS_IHD = 2.10, DIS_HYPERT = 2.03,
  CONMED_ANTIPLATELET = 0.59
)

hr_tab <- tibble::tibble(
  covariate  = names(coefs),
  paper_name = c("HPLD", "IHD", "HTN", "APLT"),
  coefficient = as.numeric(coefs),
  hr_from_model = exp(as.numeric(coefs)),
  hr_published  = as.numeric(published_hr[names(coefs)])
) |>
  dplyr::mutate(pct_diff = 100 * (hr_from_model - hr_published) / hr_published)

hr_tab |>
  dplyr::rename(
    "Covariate" = covariate,
    "Paper name" = paper_name,
    "Coefficient" = coefficient,
    "exp(coefficient)" = hr_from_model,
    "Published aHR" = hr_published,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = 3)

# All four reproduce the published adjusted hazard ratios. Three agree to
# within 0.31%; the antiplatelet row is out by 1.37% because exp(-0.514) =
# 0.5981 rounds to 0.60 rather than the 0.59 printed in Table 3 -- a rounding
# slip in the paper, not a transcription error here.
stopifnot(all(abs(hr_tab$pct_diff) < 2))
```

The same coefficients must also reproduce the hazard ratios the model itself
produces, which confirms the `model()` block applies them multiplicatively on
the hazard as Elhefnawy 2023 specifies.

```{r fig2-model-hr}
off <- dplyr::filter(sim, dplyr::if_all(dplyr::all_of(names(coefs)), ~ .x == 0))

hr_from_sim <- function(cov_name) {
  others <- setdiff(names(coefs), cov_name)
  on <- dplyr::filter(
    sim,
    .data[[cov_name]] == 1,
    dplyr::if_all(dplyr::all_of(others), ~ .x == 0)
  )
  ratio <- on$hazard_ris / off$hazard_ris[match(on$time, off$time)]
  # Proportional hazards: the ratio must be constant in time.
  stopifnot(diff(range(ratio)) < 1e-9)
  mean(ratio)
}

sim_hr <- vapply(names(coefs), hr_from_sim, numeric(1))

# The ratio must be constant in time (proportional hazards) and equal exp(beta).
stopifnot(all(abs(sim_hr - exp(coefs)) < 1e-9))
sim_hr
```

# Replicate Table 3 -- the "Half-life (Ln2/alpha)" column

```{r halflife}
hl <- tibble::tibble(
  parameter = c("alpha (<3 years)", "alpha (>=3 years)"),
  alpha = c(1.63, 0.23),
  computed = log(2) / c(1.63, 0.23),
  published = c(0.42, 3.008)
)

hl |>
  dplyr::rename(
    "Parameter" = parameter, "alpha (1/year)" = alpha,
    "ln(2)/alpha (year)" = computed, "Published (year)" = published
  ) |>
  knitr::kable(digits = 3)

stopifnot(all(abs(hl$computed - hl$published) < 0.02))
```

Note that Elhefnawy 2023 labels this column a half-life, but both alphas are
printed positive, so within each interval the hazard *doubles* on this
timescale rather than halving. The 5.06-month figure the paper gives alongside
0.42 is 0.42 years expressed in months.

# Internal identity -- closed form against numerical integration

The model file evaluates the cumulative hazard in closed form rather than by an
ODE, so that the two hazard discontinuities never reach the solver. That is
only safe if the closed form really is the integral of the instantaneous hazard
the same file reports. Integrating `hazard_ris` numerically on a fine grid must
recover `cumhaz_ris`. Both sides use the same parameter values, so this is pure
numerical error and a tight bound is the right check.

```{r identity}
# The two breakpoints are genuine jump discontinuities, so the grid carries a
# point immediately below each one. Without them the single trapezoid that
# straddles a jump averages the pre- and post-jump hazard across a full step
# and injects an error far larger than the quadrature error being measured.
fine <- sort(unique(c(seq(0, 7.37, by = 0.001),
                      0.5 - 1e-8, 0.5, 3 - 1e-8, 3, 7.37)))
fine_ev <-
  patterns |>
  dplyr::select(id, DIS_HYPERLIP, DIS_IHD, DIS_HYPERT, CONMED_ANTIPLATELET) |>
  tidyr::expand_grid(time = fine) |>
  dplyr::mutate(evid = 0, amt = 0) |>
  dplyr::arrange(id, time)

fine_sim <-
  rxode2::rxSolve(mod, fine_ev, omega = NA, sigma = NA, returnType = "data.frame") |>
  dplyr::as_tibble() |>
  dplyr::mutate(id = as.integer(as.character(id)))

trap <- function(x, y) c(0, cumsum(diff(x) * (utils::head(y, -1) + utils::tail(y, -1)) / 2))

identity_chk <-
  fine_sim |>
  dplyr::group_by(id) |>
  dplyr::arrange(time, .by_group = TRUE) |>
  dplyr::mutate(cumhaz_numeric = trap(time, hazard_ris)) |>
  dplyr::ungroup() |>
  # Relative, not absolute: the covariate multiplier scales both sides
  # identically, so an absolute bound would be ~10x tighter for the
  # covariate-free pattern than for the all-comorbidities one and would say
  # nothing about the closed form itself. Skip the first few grid points,
  # where the cumulative hazard is still near zero.
  dplyr::filter(time > 0.01) |>
  dplyr::mutate(rel_err = abs(cumhaz_numeric - cumhaz_ris) / cumhaz_ris)

max(identity_chk$rel_err)

# Composite-trapezoid relative error on exp(a t) is about (a*dx)^2/12, so a
# 0.001-year grid with a = 1.63 predicts ~2e-7. Assert an order of magnitude
# tighter than the 1e-5 that would still be comfortable, across all 16
# covariate patterns: this confirms the closed form really is the integral of
# the model's own instantaneous hazard.
stopifnot(max(identity_chk$rel_err) < 1e-5)
```

# Replicate Figure 3 -- survivor function against the published Kaplan-Meier

Figure 3 of Elhefnawy 2023 is the Kaplan-Meier VPC over the whole registry
cohort. The cohort-weighted survivor function is the weighted average of the 16
covariate patterns.

The reference values below come from two places. The value at 6 months is a
**hard published count** -- Table 1 reports 108 recurrences within 6 months
among 7,697 patients at risk. The remaining values are **read off Figure 3**
and are accurate to roughly half a percentage point; they are used only to show
the size and direction of a discrepancy that is far larger than the reading
error.

```{r fig3}
pop_surv <-
  sim |>
  dplyr::group_by(time) |>
  dplyr::summarise(sur_pop = sum(weight * sur), .groups = "drop")

published_km <- tibble::tibble(
  time = c(0.5, 1, 2, 3, 7.37),
  sur_published = c(0.977, 0.973, 0.962, 0.950, 0.918),
  provenance = c("Table 1 count (108/7,697) and Figure 3",
                 rep("read from Figure 3", 4))
)

ggplot(pop_surv, aes(time, 100 * sur_pop)) +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  geom_point(data = published_km, aes(time, 100 * sur_published),
             colour = "firebrick", size = 2) +
  coord_cartesian(ylim = c(60, 100)) +
  labs(x = "Time (years)",
       y = "Probability of not having recurrent IS (%)",
       caption = paste("Blue: cohort-weighted survivor function from the published parameters.",
                       "Red: Elhefnawy 2023 Figure 3 / Table 1.")) +
  theme_bw()
```

```{r fig3-table}
cmp <-
  published_km |>
  dplyr::mutate(
    sur_model = approx(pop_surv$time, pop_surv$sur_pop, xout = time)$y,
    cumhaz_model     = -log(sur_model),
    cumhaz_published = -log(sur_published),
    fold_over        = cumhaz_model / cumhaz_published
  )

cmp |>
  dplyr::select(time, sur_published, sur_model, cumhaz_published,
                cumhaz_model, fold_over, provenance) |>
  dplyr::rename(
    "Time (year)" = time,
    "Published S(t)" = sur_published,
    "Model S(t)" = sur_model,
    "Published cumulative hazard" = cumhaz_published,
    "Model cumulative hazard" = cumhaz_model,
    "Model / published" = fold_over,
    "Reference provenance" = provenance
  ) |>
  knitr::kable(digits = 4)
```

The model overpredicts the cumulative hazard by roughly six- to twelve-fold at
every time point once the cohort covariate distribution is applied (four- to
eight-fold for a covariate-free patient). This is asserted, not merely
displayed, so that the check fails loudly if a future edit silently changes the
transcription in either direction.

```{r fig3-assert}
stopifnot(
  # The discrepancy is real and large at every reference time.
  all(cmp$fold_over > 4),
  # ... and bounded, so an accidental order-of-magnitude edit to a theta
  # would break this rather than pass unnoticed.
  all(cmp$fold_over < 16)
)
```

## The discrepancy does not depend on how the piecewise structure is read

The reading above is the one the Table 2 footnote specifies. It is worth
showing that no alternative reading rescues the transcription, because the
first 6 months alone already falsify the scale: `theta1 = 0.238` per year is
too large by roughly an order of magnitude regardless of what the shape
parameter does.

```{r alternatives}
H_obs_6mo <- -log(1 - 108 / 7697)   # from the Table 1 count

alternatives <- tibble::tribble(
  ~reading,                                              ~H_6mo,
  "As published: h = theta1 * exp(theta2 * t)",           0.238 / 1.63 * (exp(1.63 * 0.5) - 1),
  "Constant hazard, no Gompertz factor: h = theta1",      0.238 * 0.5,
  "Decaying Gompertz: h = theta1 * exp(-theta2 * t)",     0.238 / 1.63 * (1 - exp(-1.63 * 0.5)),
  "Scale intervals swapped (theta3 first)",               0.0016 / 1.63 * (exp(1.63 * 0.5) - 1)
) |>
  dplyr::mutate(
    fold_over = H_6mo / H_obs_6mo,
    implied_6mo_incidence = 1 - exp(-H_6mo)
  )

alternatives |>
  dplyr::rename(
    "Structural reading" = reading,
    "Cumulative hazard at 6 months" = H_6mo,
    "Fold over published" = fold_over,
    "Implied 6-month incidence" = implied_6mo_incidence
  ) |>
  knitr::kable(digits = 4)
```

Against the published 108 / 7,697 = 1.40%, the as-published reading implies
17%, a flat hazard implies 11.9%, and a decaying Gompertz implies 7.8%. Only
the fourth row -- swapping the two scale parameters, which contradicts both
the Table 2 footnote and the Table 3 row labels -- lands low, and it
undershoots by an order of magnitude in the other direction. Note also that
this comparison is generous to the model: it ignores the covariate multiplier,
whose cohort average is above one and therefore makes every row worse.

```{r alternatives-assert}
cohort_multiplier <- sum(
  patterns$weight *
    exp(0.799 * patterns$DIS_HYPERLIP + 0.745 * patterns$DIS_IHD +
        0.711 * patterns$DIS_HYPERT - 0.514 * patterns$CONMED_ANTIPLATELET)
)
cohort_multiplier

stopifnot(
  # No reading consistent with the paper's own text lands near the published
  # 6-month cumulative hazard.
  all(alternatives$fold_over[1:3] > 4),
  # The cohort covariate multiplier is above 1, so it cannot close the gap.
  cohort_multiplier > 1
)
```

# Replicate Figure 4 -- survival with and without antiplatelet

Figure 4 splits the survivor function by whether the patient received an
antiplatelet at discharge. Within each stratum the remaining three covariates
are averaged over their published prevalences.

```{r fig4, fig.cap = "Replicates Figure 4 of Elhefnawy 2023 (survival with vs without antiplatelet secondary prevention). The vertical scale differs from the paper's because of the overprediction documented in the Errata; the separation between the two arms is the reproducible feature."}
by_aplt <-
  sim |>
  dplyr::group_by(CONMED_ANTIPLATELET, time) |>
  dplyr::summarise(sur_arm = sum(weight * sur) / sum(weight), .groups = "drop") |>
  dplyr::mutate(arm = ifelse(CONMED_ANTIPLATELET == 1,
                             "(ii) receiving APLT", "(i) not receiving APLT"))

ggplot(by_aplt, aes(time, 100 * sur_arm, colour = arm)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time (years)", y = "Probability of not having recurrent IS (%)",
       colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

The qualitative finding of Figure 4 -- antiplatelet-treated patients do better,
and the gap widens over time -- reproduces. For any single covariate pattern
the ratio of cumulative hazards between the two arms is exactly the published
hazard ratio, because the covariate multiplier factors out of the integral.
Averaging *survival* over the other three covariates is not the same as
averaging *hazard*, so the stratum-level ratio sits slightly above
`exp(-0.514)` and drifts upward as survival falls -- it is bounded below by the
hazard ratio and stays within a few percent of it.

```{r fig4-check}
arm_ratio <-
  by_aplt |>
  dplyr::select(CONMED_ANTIPLATELET, time, sur_arm) |>
  tidyr::pivot_wider(names_from = CONMED_ANTIPLATELET,
                     values_from = sur_arm, names_prefix = "aplt") |>
  dplyr::filter(time > 0) |>
  dplyr::mutate(cumhaz_ratio = log(aplt1) / log(aplt0))

range(arm_ratio$cumhaz_ratio)

stopifnot(
  # Bounded below by the published hazard ratio exp(-0.514) = 0.5981, which it
  # attains at early times where survival is still near 1 and the Jensen gap
  # from averaging survival rather than hazard has not yet opened.
  all(arm_ratio$cumhaz_ratio >= exp(-0.514) - 1e-6),
  # ... and never drifts more than a few percent above it.
  all(arm_ratio$cumhaz_ratio < 0.64),
  # The tightest point is within 0.1% of the published hazard ratio.
  abs(min(arm_ratio$cumhaz_ratio) - exp(-0.514)) < 1e-3
)
```

# NCA validation

Not applicable. This model has no drug-concentration output -- it produces a
hazard, a cumulative hazard and a survivor probability. PKNCA is not a
meaningful check here, so the validation above uses the checks appropriate to a
parametric time-to-event model instead: exact hazard-ratio identities against
Table 3 and Figure 2, the closed-form-versus-numerical-integration identity for
the cumulative hazard, the Table 3 half-life column, replication of Figures 1
and 4, and a quantified comparison of the survivor function against the
published Kaplan-Meier of Figure 3.

# Assumptions and deviations

* **Between-subject variability is fixed at zero.** Elhefnawy 2023 Methods
  states that "between-subject variability around the hazard was estimated,
  assuming an exponential distribution for the random effect", but Table 3
  reports no variance, no CV% and no shrinkage for it, and there is no
  supplement on disk. The structural form is therefore known and the magnitude
  is not, so `etalh0_early_ris` and `etalh0_late_ris` are declared and fixed at
  0 rather than invented. Everything in this vignette is a typical-value
  trajectory. The paper describes a single random effect on the hazard; two are
  declared here only so that each pairs with a fixed-effect parameter, and
  anyone reinstating the variability should draw one value and use it for both.
* **`addSd` is not from the source.** The source fits the parametric survival
  (event-density) likelihood under LAPLACE, so there is no observation-error
  model to translate. A placeholder additive residual of 0.001 is attached to
  `sur` so that the nlmixr2 likelihood machinery accepts the model for forward
  simulation.
* **Covariate independence.** The 16-pattern weights assume the four covariates
  are independent, because Elhefnawy 2023 does not publish their joint
  distribution. The paper's own Limitations section says "the comorbidities
  were analyzed independently", so this matches how the model was built, but it
  is an assumption: hypertension, hyperlipidemia and ischemic heart disease are
  positively correlated in real cohorts, which would concentrate risk in fewer
  patients than assumed here.
* **The Figure 3 reference points at 1, 2, 3 and 7.37 years are digitized**
  from the published figure and are accurate to roughly half a percentage
  point. The 6-month value is a hard count from Table 1. The conclusions drawn
  from them concern a four- to eight-fold discrepancy, which is far outside the
  digitization error.
* **The cumulative hazard is evaluated in closed form, not by an ODE.** This is
  an encoding choice, not a deviation from the published model: the piecewise
  Gompertz integrates elementally, and integrating it analytically keeps the
  two hazard discontinuities out of the solver, where they would be stepped
  over and silently smeared. The closed form is verified against numerical
  integration of the model's own instantaneous hazard above.
* **Antiplatelet class composition is unknown.** The paper does not enumerate
  which agents were pooled into the antiplatelet indicator, and the flag records
  a prescription written at discharge rather than measured adherence.
* **Hypertension and diabetes duration are not used.** Table 1 stratifies both
  by duration, but only the presence/absence flags appear in the final model.

# Errata

**The published parameters do not reproduce the paper's own outputs.** This is
recorded here in full because it is material to anyone using the model, and
because nothing has been adjusted to hide it.

The baseline-hazard scale for the first 6 months, `theta1 = 0.238` per year, is
irreconcilable with the paper's own event counts:

* Elhefnawy 2023 Table 1 reports 108 recurrences within 6 months among 7,697
  patients at risk, a cumulative hazard of about 0.0141.
* As published, `h(t) = 0.238 * exp(1.63 t)` integrates to 0.184 over the first
  6 months -- 13-fold too high, implying a 17% 6-month recurrence rate against
  the observed 1.40%.
* The discrepancy is not an artefact of the Gompertz shape. A merely *constant*
  hazard of 0.238 per year gives 0.119 over 6 months, still 8.5-fold too high.
  The scale itself is too large.
* The covariate multiplier cannot rescue it: averaged over the published
  prevalences it is about 1.66, which makes the overprediction worse rather
  than better.
* Between-subject variability cannot rescue it either. A zero-mean random
  effect on the log hazard leaves at least half the cohort at or above the
  typical hazard, so the population survivor function is bounded well below the
  published curve no matter how large the unreported variance is.

The same overprediction persists across the whole follow-up when compared with
the Figure 3 Kaplan-Meier VPC. For a covariate-free patient the model's
cumulative hazard runs about 8-fold high at 6 months, 6-fold at 3 years and
4-fold at 7.37 years; applying the cohort covariate distribution raises those
to roughly 12-fold, 9-fold and 6-fold. The `Replicate Figure 3` section above
asserts this numerically.

Two smaller inconsistencies are noted for completeness:

* **The sign of `theta4`.** Results states that the hazard showed "the
  exponential increase ... in the first 3 years after the index IS and then
  exponentially reduced afterward". With `theta4 = +0.23` as printed, the
  hazard still increases after 3 years, only about seven times more slowly. The
  "reduction" the sentence describes is produced by the downward jump at
  `t = 3` that the scale/shape switching creates, not by a negative exponent.
  The printed positive sign is what is encoded.
* **The Table 3 "Half-life (Ln2/alpha)" column.** `ln(2)/alpha` is a doubling
  time for a positive exponent, not a half-life. The arithmetic reproduces
  (0.425 and 3.014 years against the published 0.42 and 3.008); only the label
  is wrong.

What *does* reproduce exactly is the entire covariate block: all four
coefficients back out to the published adjusted hazard ratios to within the
paper's own two-decimal rounding, and the relative effects in Figures 2 and 4
are reproduced faithfully. The model is therefore usable for *relative* risk
questions -- how much does hyperlipidemia or an antiplatelet prescription move
the hazard -- while its absolute recurrence predictions should not be used
without resolving the baseline-hazard scale with the authors.

No supplementary material was available on disk for this paper; the Frontiers
supplementary-material link is referenced in the article but the file was not
part of the open-access package retrieved for this extraction. A supplement
containing the NONMEM control stream would very likely settle the scale
question.
