---
title: "Recurrent ischemic stroke with and without diabetes (Elhefnawy 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recurrent ischemic stroke with and without diabetes (Elhefnawy 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Elhefnawy M, Noor Harun S, Leykhim T, Tangiisuran B, Zainal H,
  Looi I, Sidek N, Abdul Aziz Z, Sheikh Ghadzi SM. A parametric time-to-event
  modelling of recurrent ischemic stroke after index stroke among patients with
  and without diabetes mellitus: implementation of temporal validation of the
  model. Cureus. 2023;15(12):e50794. doi:10.7759/cureus.50794.
* Article: <https://doi.org/10.7759/cureus.50794> (Open Access, CC BY 4.0)
* Model files:
  `modellib("Elhefnawy_2023_recurrent_ischemic_stroke_dm")` and
  `modellib("Elhefnawy_2023_recurrent_ischemic_stroke_nondm")`

The paper develops **two** parametric time-to-event (TTE) models of recurrent
ischemic stroke after a first (index) ischemic stroke, one in each stratum of
the National Neurology Registry of Malaysia split by diabetes status. The two
models share a structure but have separate estimates and separate retained
covariates, so they are packaged as two model files and documented here in one
vignette.

Neither is a PK or PK-PD model: there is no drug exposure term. Secondary
prevention enters the non-diabetic model only as a binary "antihyperlipidemic
prescribed" indicator. Time is in years throughout.

**Read the Errata section before using either model.** The published
baseline-hazard scale does not reproduce the paper's own event counts, and the
paper's own clinical calculator does not reproduce its own parameters. The
arithmetic is set out in full below. Everything in these files is transcribed
exactly as published; nothing has been tuned.

```{r model}
mod_dm <- modellib("Elhefnawy_2023_recurrent_ischemic_stroke_dm")
mod_nd <- modellib("Elhefnawy_2023_recurrent_ischemic_stroke_nondm")
mod_dm
```

# Population

National Neurology Registry (NNEUR) of Malaysia, a hospital-based registry
covering 13 states. Patients with a first ischemic stroke registered between
August 2009 and December 2016 were extracted and stratified by diabetes status.
The endpoint is any subsequent ischemic stroke recorded in the registry;
follow-up was censored at the last recorded observation, to a maximum of 7.37
years. Diabetes was identified from physician diagnosis, antidiabetic
medication history, the electronic record, or antidiabetic medications
prescribed at discharge (Methods, "Data collection").

The diabetic stratum was additionally validated **temporally** on a separate
NNEUR cohort of 1,262 diabetic patients with index strokes registered January
2017 to December 2020 (Methods, "Temporal validation"; Figure 4). No parameter
estimates are reported for that cohort, so it is described but not encoded.

```{r population}
pop <- data.frame(
  Quantity = c(
    "Patients", "Recurrent ischemic strokes", "Recurrence (%)",
    "Maximum follow-up (years)",
    "Hypertension (%)", "Hyperlipidemia (%)", "Ischemic heart disease (%)",
    "Atrial fibrillation (%)", "Hyperuricemia (%)",
    "Current smoker (%)", "Female (%)",
    "Antihyperlipidemic prescribed (%)", "Antiplatelet prescribed (%)"
  ),
  `Diabetic stratum` = c(
    3493, 195, round(100 * 195 / 3493, 2), 7.37,
    round(100 * (180 + 2863) / 3493, 2), round(100 * (96 + 1004) / 3493, 2),
    round(100 * (52 + 420) / 3493, 2), round(100 * (4 + 87) / 3493, 2),
    round(100 * (10 + 121) / 3493, 2), round(100 * (113 + 1630) / 3493, 2),
    round(100 * (101 + 1647) / 3493, 2), round(100 * (167 + 2926) / 3493, 2),
    round(100 * (167 + 2978) / 3493, 2)
  ),
  `Non-diabetic stratum` = c(
    4204, 138, round(100 * 138 / 4204, 2), 7.37,
    round(100 * (108 + 2355) / 4204, 2), round(100 * (63 + 865) / 4204, 2),
    round(100 * (25 + 382) / 4204, 2), round(100 * (5 + 172) / 4204, 2),
    round(100 * (6 + 97) / 4204, 2), round(100 * (89 + 1917) / 4204, 2),
    round(100 * (53 + 1607) / 4204, 2), round(100 * (120 + 3682) / 4204, 2),
    round(100 * (118 + 3635) / 4204, 2)
  ),
  check.names = FALSE
)
knitr::kable(pop, caption = "Cohort composition, computed from the recurrent and non-recurrent counts of Elhefnawy 2023 Table 1.")
```

The paper's own recurrence percentages disagree with its counts in the diabetic
stratum: the Abstract prints 5.82%, the Results section prints 5.55%, and
195 / 3,493 is 5.58%. The counts are used throughout this vignette.

```{r population-check}
stopifnot(
  abs(100 * 195 / 3493 - 5.58) < 0.01,
  abs(100 * 138 / 4204 - 3.28) < 0.01
)
```

# Source trace

```{r source-trace}
trace <- tibble::tribble(
  ~Item, ~`Source location`, ~Value,
  "S(t) = exp(-integral h)", "Equation 1 (Methods, 'Base Model Development')", "structure",
  "h(t) = h0 * exp(theta_y * t)", "Table 2 row 4 ('After inserting different time intervals', Gompertz)", "structure",
  "h0 = theta1 if t < 0.5 y, theta3 if t >= 0.5 y", "Table 2 footnote; restated in Methods", "structure",
  "theta_y = theta2 if t < 3 y, theta4 if t >= 3 y", "Table 2 footnote; restated in Methods", "structure",
  "h = h0 * exp(beta_t*t + beta_cov*(cov))", "Equation 5, printed inline on page 2 of the PDF", "structure",
  "theta1 (DM)", "Table 3 row 1, DM column", "0.356 /y (RSE 13.69%)",
  "theta3 (DM)", "Table 3 row 2, DM column", "0.0023 /y (RSE 17.01%)",
  "theta2 (DM)", "Table 3 row 3, DM column", "1.58 /y (RSE 5.98%)",
  "theta4 (DM)", "Table 3 row 4, DM column", "0.242 /y (RSE 22.36%)",
  "IHD effect (DM)", "Table 3 row 5, DM column", "0.876 (aHR 2.40, 1.79-3.20)",
  "HPLD effect (DM)", "Table 3 row 6, DM column", "0.633 (aHR 1.88, 1.44-2.45)",
  "theta1 (non-DM)", "Table 3 row 1, non-DM column", "0.253 /y (RSE 24.38%)",
  "theta3 (non-DM)", "Table 3 row 2, non-DM column", "0.0018 /y (RSE 23.09%)",
  "theta2 (non-DM)", "Table 3 row 3, non-DM column", "1.7 /y (RSE 6.37%)",
  "theta4 (non-DM)", "Table 3 row 4, non-DM column", "0.213 /y (RSE 33.07%)",
  "HPLD effect (non-DM)", "Table 3 row 6, non-DM column", "1.03 (aHR 2.801, 2.00-3.90)",
  "HTN effect (non-DM)", "Table 3 row 7, non-DM column", "0.789 (aHR 2.201, 1.53-3.14)",
  "Antihyperlipidemic effect (non-DM)", "Table 3 row 8, non-DM column", "-0.835 (aHR 0.433, 0.65-0.28)",
  "Breakpoints 0.5 y and 3 y", "Table 2 footnote; not estimated (Table 2 counts 4 parameters)", "fixed",
  "Between-subject variability", "not reported anywhere in the paper", "absent",
  "Residual error", "n/a -- survival likelihood, LAPLACE (Methods, 'Model development')", "placeholder"
)
knitr::kable(trace, caption = "Provenance of every structural equation and every ini() value.")
```

# Structural behaviour

## Baseline hazard: replicating Figure 1

Figure 1 of Elhefnawy 2023 contrasts the baseline hazard during and after the
first six months in each stratum. The Abstract and Results give the numbers
directly: 0.356 and 0.253 within the first six months, 0.0023 and 0.0018
afterwards.

```{r figure1}
baseline <- tibble::tribble(
  ~Stratum,        ~Interval,          ~`Baseline hazard (1/year)`,
  "Diabetes",      "< 6 months",       0.356,
  "Diabetes",      ">= 6 months",      0.0023,
  "No diabetes",   "< 6 months",       0.253,
  "No diabetes",   ">= 6 months",      0.0018
)
ggplot(baseline, aes(Interval, `Baseline hazard (1/year)`, fill = Stratum)) +
  geom_col(position = position_dodge()) +
  scale_y_log10() +
  labs(
    title = "Replicates Figure 1 of Elhefnawy 2023",
    subtitle = "Baseline hazard scale, log axis",
    y = "Baseline hazard (1/year, log scale)"
  ) +
  theme_bw()
```

## The hazard trajectory and its two discontinuities

The scale switches at 0.5 year and the shape switches independently at 3 years,
so the hazard has **three** regimes and **two** downward jumps. The
`hazard_ris` output of each model makes them visible.

```{r hazard-traj}
grid <- seq(0, 7.37, length.out = 1475)

solve_pattern <- function(mod, covs, times = grid) {
  d <- as.data.frame(rxode2::et(times))
  for (nm in names(covs)) d[[nm]] <- covs[[nm]]
  rxode2::rxSolve(mod, d, returnType = "data.frame")
}

haz <- bind_rows(
  solve_pattern(mod_dm, list(DIS_IHD = 0, DIS_HYPERLIP = 0)) |>
    transmute(time, hazard = hazard_ris, cumhaz = cumhaz_ris, sur, Stratum = "Diabetes"),
  solve_pattern(mod_nd, list(DIS_HYPERLIP = 0, DIS_HYPERT = 0, CONMED_LIPIDLOWER = 0)) |>
    transmute(time, hazard = hazard_ris, cumhaz = cumhaz_ris, sur, Stratum = "No diabetes")
)

ggplot(haz, aes(time, hazard, colour = Stratum)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  geom_vline(xintercept = c(0.5, 3), linetype = "dashed", colour = "grey50") +
  labs(
    title = "Baseline hazard of recurrent ischemic stroke, covariate-free patient",
    subtitle = "Dashed lines: the scale switch at 0.5 y and the shape switch at 3 y",
    x = "Years after the index stroke", y = "Hazard (1/year, log scale)"
  ) +
  theme_bw()
```

The jump sizes follow directly from the parameters and are asserted here so a
later edit cannot silently change the structure.

```{r discontinuities}
jump <- tibble::tibble(
  Stratum = c("Diabetes", "No diabetes"),
  `Drop at 0.5 y (fold)` = c(0.356 / 0.0023, 0.253 / 0.0018),
  `Drop at 3 y (fold)` = c(exp(1.58 * 3) / exp(0.242 * 3), exp(1.7 * 3) / exp(0.213 * 3))
)
knitr::kable(jump, digits = 1)

stopifnot(
  abs(0.356 / 0.0023 - 154.8) < 0.1,
  abs(0.253 / 0.0018 - 140.6) < 0.1,
  abs(exp(1.58 * 3) / exp(0.242 * 3) - 55.4) < 0.1,
  abs(exp(1.7 * 3) / exp(0.213 * 3) - 86.6) < 0.1
)
```

Note the direction. Elhefnawy 2023 Results says the hazard "increased
exponentially during the first three years after the index IS and then
exponentially reduced afterwards", but both late shape parameters are printed
**positive** (0.242 and 0.213), so within the third regime the hazard still
rises -- roughly seven-fold more slowly. The apparent reduction is the jump at
t = 3 produced by the shape switch, not a negative exponent. The printed sign is
encoded; see Errata.

## Internal identity: the closed-form cumulative hazard

The model evaluates the cumulative hazard analytically rather than by an ODE,
which keeps the two discontinuities out of the solver. That closed form is
checked here against trapezoidal integration of the model's own `hazard_ris`
output, over every covariate pattern in both models. Both sides use the same
drawn parameters, so this is pure numerical error and a tight bound is
appropriate.

```{r closed-form-check}
fine <- seq(0, 7.37, length.out = 20001)

max_rel_err <- function(mod, covs) {
  s <- solve_pattern(mod, covs, fine)
  trap <- cumsum(c(0, diff(s$time) *
                     (head(s$hazard_ris, -1) + tail(s$hazard_ris, -1)) / 2))
  max(abs(trap - s$cumhaz_ris) / pmax(s$cumhaz_ris, 1e-12))
}

dm_patterns <- expand.grid(DIS_IHD = 0:1, DIS_HYPERLIP = 0:1)
nd_patterns <- expand.grid(DIS_HYPERLIP = 0:1, DIS_HYPERT = 0:1, CONMED_LIPIDLOWER = 0:1)

dm_err <- vapply(seq_len(nrow(dm_patterns)),
                 function(i) max_rel_err(mod_dm, as.list(dm_patterns[i, ])), numeric(1))
nd_err <- vapply(seq_len(nrow(nd_patterns)),
                 function(i) max_rel_err(mod_nd, as.list(nd_patterns[i, ])), numeric(1))

cat(sprintf("Diabetic model:     %d covariate patterns, max relative error %.2e\n",
            nrow(dm_patterns), max(dm_err)))
cat(sprintf("Non-diabetic model: %d covariate patterns, max relative error %.2e\n",
            nrow(nd_patterns), max(nd_err)))

stopifnot(max(dm_err) < 1e-3, max(nd_err) < 1e-3)
```

# Reproducing the paper's published quantities

## Adjusted hazard ratios (Table 3, Figure 2)

Every retained covariate is a 0/1 indicator, so the adjusted hazard ratio is
`exp()` of the coefficient. All five reproduce the printed values.

```{r ahr}
ahr <- tibble::tribble(
  ~Model, ~Covariate, ~Coefficient, ~`Published aHR`,
  "Diabetes",    "Ischemic heart disease",  0.876,  2.40,
  "Diabetes",    "Hyperlipidemia",          0.633,  1.88,
  "No diabetes", "Hyperlipidemia",          1.030,  2.801,
  "No diabetes", "Hypertension",            0.789,  2.201,
  "No diabetes", "Antihyperlipidemic",     -0.835,  0.433
) |>
  mutate(
    `Recomputed aHR` = exp(Coefficient),
    `Difference (%)` = 100 * (`Recomputed aHR` - `Published aHR`) / `Published aHR`
  )
knitr::kable(ahr, digits = 4,
             caption = "exp(coefficient) against the adjusted hazard ratios of Elhefnawy 2023 Table 3.")

stopifnot(max(abs(ahr$`Difference (%)`)) < 0.25)
```

The model files must produce the same ratios when solved, which is the check
that the coefficients were wired into the right covariate columns.

```{r ahr-from-solve}
haz_at <- function(mod, covs, tt = 1) {
  solve_pattern(mod, covs, c(0, tt))$hazard_ris[2]
}
solved_ahr <- c(
  IHD_dm  = haz_at(mod_dm, list(DIS_IHD = 1, DIS_HYPERLIP = 0)) /
            haz_at(mod_dm, list(DIS_IHD = 0, DIS_HYPERLIP = 0)),
  HPLD_dm = haz_at(mod_dm, list(DIS_IHD = 0, DIS_HYPERLIP = 1)) /
            haz_at(mod_dm, list(DIS_IHD = 0, DIS_HYPERLIP = 0)),
  HPLD_nd = haz_at(mod_nd, list(DIS_HYPERLIP = 1, DIS_HYPERT = 0, CONMED_LIPIDLOWER = 0)) /
            haz_at(mod_nd, list(DIS_HYPERLIP = 0, DIS_HYPERT = 0, CONMED_LIPIDLOWER = 0)),
  HTN_nd  = haz_at(mod_nd, list(DIS_HYPERLIP = 0, DIS_HYPERT = 1, CONMED_LIPIDLOWER = 0)) /
            haz_at(mod_nd, list(DIS_HYPERLIP = 0, DIS_HYPERT = 0, CONMED_LIPIDLOWER = 0)),
  LL_nd   = haz_at(mod_nd, list(DIS_HYPERLIP = 0, DIS_HYPERT = 0, CONMED_LIPIDLOWER = 1)) /
            haz_at(mod_nd, list(DIS_HYPERLIP = 0, DIS_HYPERT = 0, CONMED_LIPIDLOWER = 0))
)
print(round(solved_ahr, 4))

stopifnot(max(abs(solved_ahr - c(2.40, 1.88, 2.801, 2.201, 0.433)) /
                c(2.40, 1.88, 2.801, 2.201, 0.433)) < 0.0025)
```

## The Table 3 half-life column

Table 3 carries a column labelled "Half-life (Ln2/alpha)" for each shape
parameter. The arithmetic reproduces exactly for all four.

```{r halflife}
hl <- tibble::tribble(
  ~Model, ~Shape, ~alpha, ~`Published (paper text)`,
  "Diabetes",    "alpha (<3)",  1.58,  "0.43 (5.25 months)",
  "Diabetes",    "alpha (>=3)", 0.242, "2.85 years",
  "No diabetes", "alpha (<3)",  1.70,  "0.40 (4.88 months)",
  "No diabetes", "alpha (>=3)", 0.213, "3.24 years"
) |>
  mutate(
    `Recomputed (years)` = log(2) / alpha,
    `Recomputed (months)` = 12 * log(2) / alpha
  )
knitr::kable(hl, digits = 3)

stopifnot(
  abs(log(2) / 1.58  - 0.4387) < 1e-3,
  abs(log(2) / 0.242 - 2.8642) < 1e-3,
  abs(log(2) / 1.70  - 0.4077) < 1e-3,
  abs(log(2) / 0.213 - 3.2542) < 1e-3,
  abs(12 * log(2) / 1.58 - 5.25) < 0.02,
  abs(12 * log(2) / 1.70 - 4.88) < 0.02
)
```

A naming caveat worth recording: for a **positive** exponent, `ln(2)/alpha` is a
**doubling** time, not a half-life. The arithmetic is right; the column label is
not.

# Deterministic cohort simulation

The paper reports no between-subject variability, so there is nothing to draw
from and no reason to run a Monte Carlo cohort. Instead every covariate pattern
is enumerated and weighted by its Table 1 prevalence (assuming independence
between the indicators, which is all the marginal table supports). This gives
the population-average recurrence-free survivor curve exactly and
reproducibly -- there is no random number generation anywhere in this vignette.

```{r cohort}
p_ihd_dm  <- (52 + 420)  / 3493
p_hpld_dm <- (96 + 1004) / 3493
p_hpld_nd <- (63 + 865)  / 4204
p_htn_nd  <- (108 + 2355) / 4204
p_ll_nd   <- (120 + 3682) / 4204

wt <- function(x, p) ifelse(x == 1, p, 1 - p)

dm_cohort <- dm_patterns |>
  mutate(weight = wt(DIS_IHD, p_ihd_dm) * wt(DIS_HYPERLIP, p_hpld_dm))
nd_cohort <- nd_patterns |>
  mutate(weight = wt(DIS_HYPERLIP, p_hpld_nd) * wt(DIS_HYPERT, p_htn_nd) *
           wt(CONMED_LIPIDLOWER, p_ll_nd))

stopifnot(abs(sum(dm_cohort$weight) - 1) < 1e-12,
          abs(sum(nd_cohort$weight) - 1) < 1e-12)

avg_survivor <- function(mod, cohort, covnames) {
  parts <- lapply(seq_len(nrow(cohort)), function(i) {
    s <- solve_pattern(mod, as.list(cohort[i, covnames]))
    s$sur * cohort$weight[i]
  })
  tibble::tibble(time = grid, sur = Reduce(`+`, parts))
}

surv <- bind_rows(
  avg_survivor(mod_dm, dm_cohort, c("DIS_IHD", "DIS_HYPERLIP")) |>
    mutate(Stratum = "Diabetes"),
  avg_survivor(mod_nd, nd_cohort, c("DIS_HYPERLIP", "DIS_HYPERT", "CONMED_LIPIDLOWER")) |>
    mutate(Stratum = "No diabetes")
)

ggplot(surv, aes(time, sur, colour = Stratum)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = c(1 - 195 / 3493, 1 - 138 / 4204) |> rev(),
             linetype = "dotted", colour = "grey40") +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "Model-predicted probability of remaining free of recurrent stroke",
    subtitle = "Compare Figure 3 of Elhefnawy 2023. Dotted lines: the observed recurrence-free fraction at 7.37 y",
    x = "Years after the index stroke", y = "Recurrence-free probability"
  ) +
  theme_bw()
```

The paper's qualitative claim -- that recurrence-free survival is lower among
patients with diabetes -- is reproduced by the packaged models. That claim
holds throughout follow-up and is asserted here.

```{r qualitative-claim}
wide <- surv |> pivot_wider(names_from = Stratum, values_from = sur)
stopifnot(all(wide$Diabetes <= wide$`No diabetes` + 1e-12))
cat(sprintf("Predicted recurrence-free probability at 7.37 y: DM %.3f, non-DM %.3f\n",
            tail(wide$Diabetes, 1), tail(wide$`No diabetes`, 1)))
```

# Errata

Read this section before using either model. It is asserted numerically so that
a later edit cannot quietly remove it.

## 1. The baseline-hazard scale overpredicts the paper's own event count by roughly an order of magnitude

This is the load-bearing defect. Elhefnawy 2023 Table 1 reports 195 recurrences
among 3,493 diabetic patients and 138 among 4,204 non-diabetic patients over at
most 7.37 years. Those counts pin the observed cumulative hazard. The published
parameters give something far larger.

```{r falsifier-main}
obs_H <- c(Diabetes = -log(1 - 195 / 3493), `No diabetes` = -log(1 - 138 / 4204))

pred_H_baseline <- c(
  Diabetes      = tail(solve_pattern(mod_dm, list(DIS_IHD = 0, DIS_HYPERLIP = 0))$cumhaz_ris, 1),
  `No diabetes` = tail(solve_pattern(mod_nd, list(DIS_HYPERLIP = 0, DIS_HYPERT = 0,
                                                  CONMED_LIPIDLOWER = 0))$cumhaz_ris, 1)
)

# Cohort-average covariate multiplier under the Table 1 marginal prevalences.
mult_dm <- sum(dm_cohort$weight * exp(0.876 * dm_cohort$DIS_IHD +
                                        0.633 * dm_cohort$DIS_HYPERLIP))
mult_nd <- sum(nd_cohort$weight * exp(1.030 * nd_cohort$DIS_HYPERLIP +
                                        0.789 * nd_cohort$DIS_HYPERT +
                                        (-0.835) * nd_cohort$CONMED_LIPIDLOWER))

fals <- tibble::tibble(
  Stratum = names(obs_H),
  `Observed H(7.37)` = as.numeric(obs_H),
  `Predicted H(7.37), covariate-free` = as.numeric(pred_H_baseline),
  `Fold (covariate-free)` = as.numeric(pred_H_baseline / obs_H),
  `Cohort covariate multiplier` = c(mult_dm, mult_nd),
  `Fold (cohort-weighted)` = as.numeric(pred_H_baseline * c(mult_dm, mult_nd) / obs_H)
)
knitr::kable(fals, digits = 4)

stopifnot(
  fals$`Fold (covariate-free)`[1] > 7, fals$`Fold (covariate-free)`[1] < 10,
  fals$`Fold (covariate-free)`[2] > 10, fals$`Fold (covariate-free)`[2] < 14,
  fals$`Fold (cohort-weighted)`[1] > 11,
  fals$`Fold (cohort-weighted)`[2] > 12
)
```

Applying the covariate distribution makes it worse, not better, because both
cohort multipliers exceed one.

## 2. No alternative reading of the structure rescues it

Four candidate re-readings were evaluated against the same observed cumulative
hazard. All of them fail, and the constant-scale row is decisive: it removes the
Gompertz shape entirely and is still several-fold too high, which localises the
defect to the **scale** rather than to any misreading of the piecewise
structure.

```{r falsifier-alternatives}
cum_closed <- function(t, h1, h3, a2, a4, tb1 = 0.5, tb2 = 3) {
  seg <- function(h, a, lo, hi) h / a * (exp(a * hi) - exp(a * lo))
  s <- seg(h1, a2, 0, min(t, tb1))
  if (t > tb1) s <- s + seg(h3, a2, tb1, min(t, tb2))
  if (t > tb2) s <- s + seg(h3, a4, tb2, t)
  s
}
tmax <- 7.37
alt <- tibble::tibble(
  Reading = c("As published", "Constant scale, no Gompertz shape",
              "Decaying Gompertz (shapes negated)", "The two scales swapped",
              "Observed (Table 1 counts)"),
  Diabetes = c(
    cum_closed(tmax, 0.356, 0.0023, 1.58, 0.242),
    0.356 * 0.5 + 0.0023 * (tmax - 0.5),
    cum_closed(tmax, 0.356, 0.0023, -1.58, -0.242),
    cum_closed(tmax, 0.0023, 0.356, 1.58, 0.242),
    as.numeric(obs_H[1])
  ),
  `No diabetes` = c(
    cum_closed(tmax, 0.253, 0.0018, 1.70, 0.213),
    0.253 * 0.5 + 0.0018 * (tmax - 0.5),
    cum_closed(tmax, 0.253, 0.0018, -1.70, -0.213),
    cum_closed(tmax, 0.0018, 0.253, 1.70, 0.213),
    as.numeric(obs_H[2])
  )
) |>
  mutate(
    `Fold, DM` = Diabetes / as.numeric(obs_H[1]),
    `Fold, non-DM` = `No diabetes` / as.numeric(obs_H[2])
  )
knitr::kable(alt, digits = 4,
             caption = "Cumulative hazard at 7.37 years under four readings of the published structure.")

# Every alternative reading remains at least twice the observed cumulative
# hazard; none is a rescue.
stopifnot(all(alt$`Fold, DM`[1:4] > 2), all(alt$`Fold, non-DM`[1:4] > 2))

# The first six months alone, with the shape removed, already exceed the whole
# observed follow-up hazard several-fold.
stopifnot(
  0.356 * 0.5 / as.numeric(obs_H[1]) > 3,
  0.253 * 0.5 / as.numeric(obs_H[2]) > 3
)
```

A frailty term cannot rescue it either, and no unreported between-subject
variability is available to invoke: a zero-mean random effect on a log hazard
leaves at least half the cohort at or above the typical hazard, which bounds the
population survivor function well below the published Kaplan-Meier curve however
large the variance.

## 3. The paper's own clinical calculator does not reproduce its own parameters, and is internally inconsistent

Results, "Clinical Application", gives two scenarios from the authors'
`MyReCuRIS_DM` calculator. Neither matches the Table 3 parameters, and scenario
1 is impossible on its face: a cumulative recurrence probability cannot fall
over time.

```{r falsifier-calculator}
prob <- function(mod, covs, t) 1 - solve_pattern(mod, covs, c(0, t))$sur[2]
calc <- tibble::tibble(
  Scenario = c("1: DM, no HPLD, no IHD", "1: DM, no HPLD, no IHD",
               "2: DM with IHD", "2: DM with IHD"),
  `Years after index stroke` = c(1, 4, 1, 4),
  `Published (%)` = c(8.396, 4.917, 11.403, 18.989),
  `Model (%)` = 100 * c(
    prob(mod_dm, list(DIS_IHD = 0, DIS_HYPERLIP = 0), 1),
    prob(mod_dm, list(DIS_IHD = 0, DIS_HYPERLIP = 0), 4),
    prob(mod_dm, list(DIS_IHD = 1, DIS_HYPERLIP = 0), 1),
    prob(mod_dm, list(DIS_IHD = 1, DIS_HYPERLIP = 0), 4)
  )
)
knitr::kable(calc, digits = 3)

# Scenario 1 falls from 1 y to 4 y, which no cumulative probability can do.
stopifnot(calc$`Published (%)`[2] < calc$`Published (%)`[1])
# The model's own predictions are monotone increasing, as they must be.
stopifnot(calc$`Model (%)`[2] > calc$`Model (%)`[1],
          calc$`Model (%)`[4] > calc$`Model (%)`[3])
# The published scenario ratio does not equal the published IHD hazard ratio at
# either time point.
ratio_pub <- calc$`Published (%)`[c(3, 4)] / calc$`Published (%)`[c(1, 2)]
stopifnot(all(abs(ratio_pub - 2.40) > 0.5))
```

## 4. The sign of the late shape parameter contradicts the Results prose

Results says the hazard "exponentially reduced" after three years, but theta4 is
printed positive in both strata (0.242 and 0.213), so the hazard continues to
rise within the third regime. The reduction is the discontinuous drop at t = 3
produced by the shape switch. The printed sign is encoded; only the prose is
loose.

## 5. Minor: the diabetic-stratum recurrence percentage is printed two different ways

The Abstract says 5.82%, the Results section says 5.55%, and 195 / 3,493 is
5.58%. The counts are used.

## 6. The same defect is present in this group's companion publication

The pooled-cohort model published by the same group (Elhefnawy 2023, Front
Neurol 14:1118711, PMC10176964, packaged as
`Elhefnawy_2023_recurrent_ischemic_stroke`) exhibits the same overprediction at
the same order of magnitude (8.4- to 13-fold). That the defect recurs across two
independent fits to the same registry, while the covariate block reproduces
exactly in both, points to a systematic issue with how the baseline-hazard scale
is reported rather than to a transcription error in either extraction.

# Assumptions and deviations

* **Values are transcribed verbatim and untuned.** No parameter was adjusted to
  improve agreement with any published output. The discrepancies above are the
  publication's, not this extraction's.
* **These models are usable for relative risk, not absolute risk.** The hazard
  ratios reproduce exactly and are the trustworthy part; the absolute recurrence
  probabilities are roughly an order of magnitude too high.
* **No between-subject variability.** The paper reports no random effect for
  either stratum -- no variance, no CV%, no shrinkage, no omega row. The models
  are encoded faithfully as typical-value models with no eta parameters. No
  variance was invented.
* **Residual error is a placeholder.** The source fits a parametric survival
  likelihood under LAPLACE, so there is no observation-error model to translate.
  A tiny additive residual (`addSd = 0.001`) is attached to the survivor output
  so the nlmixr2 machinery accepts the model for forward simulation. It is not
  from the source.
* **Breakpoints are fixed, not estimated.** Table 2 counts the interval Gompertz
  model as having four parameters (theta1-theta4), so the 0.5-year and 3-year
  breakpoints carry no degrees of freedom.
* **Covariate independence in the cohort weighting.** Table 1 gives only
  marginal prevalences, so the enumerated cohort above assumes the indicators
  are independent. The real registry almost certainly has positive correlation
  between hypertension, hyperlipidemia and ischemic heart disease, which would
  change the cohort-average multiplier somewhat. It would not change the
  conclusion of the Errata section, whose covariate-free rows are unaffected.
* **Equations 1-4 are images in the source PDF and did not extract as text.**
  The Gompertz hazard form was read from Table 2 (which prints
  `h(t) = h0 * e^(theta_y)t` in the model column), the piecewise switching from
  the Table 2 footnote and the Methods restatement, and the covariate form from
  the Equation 5 fragment that appears inline in the page-2 text layer,
  `h = h0 * exp(beta_t + beta_cov(cov))`, read together with the Methods
  sentence "where h0 is the baseline hazard, and beta_t is the shape parameter
  of the Gompertz hazard function". These readings are mutually consistent and
  are the only ones that reproduce the Table 3 half-life and aHR columns.
* **The covariate screen is not in this paper.** Methods defers it to "a
  previously published preprint", and the EuropePMC open-access deposit for
  PMC10796130 contains no supplementary file. Screened-but-not-retained
  covariates are therefore documented in each model file's
  `covariatesDataExcluded` with prevalences from Table 1 but with no
  objective-function changes.
* **The temporal-validation cohort is not encoded.** The 1,262-patient
  2017-2020 diabetic cohort has no separate parameter estimates in the paper.
* **`CONMED_LIPIDLOWER` class composition is unknown.** The paper tabulates
  "Antihyperlipidemic" without enumerating the agents. The Discussion argues the
  effect in terms of statins, but the covariate as fitted is the broader class
  indicator, so the broader canonical is used.
