---
title: "Whole-body PBPK of intravenous iron in mice, rats and humans (Fan 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Whole-body PBPK of intravenous iron in mice, rats and humans (Fan 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7.2, fig.height = 5.4, dpi = 96
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Fan et al. built a 14-compartment perfusion-limited whole-body PBPK model of
iron in the mouse, fitted it simultaneously to `r 13` measured compartments
under three dietary iron statuses, and then extrapolated it to the rat and to
humans for the iron-carbohydrate complex preparation (IVIP) ferric
carboxymaltose (FCM).

* Article: <https://doi.org/10.1007/s13346-024-01675-x> (open access)
* Supplement (MOESM1): rat and human physiology tables, the **complete NONMEM
  control stream** for the mouse run, and every digitised observed dataset.

The paper contributed five model files, all pointing at this one vignette:

```{r models}
models <- c(
  "Fan_2025_iron_mouse_irondeficient_pbpk",
  "Fan_2025_iron_mouse_ironadequate_pbpk",
  "Fan_2025_iron_mouse_ironloaded_pbpk",
  "Fan_2025_iron_ferriccarboxymaltose_rat_pbpk",
  "Fan_2025_iron_ferriccarboxymaltose_human_pbpk"
)
# readModelDb() returns the model FUNCTION; it must be wrapped in
# rxode2::rxode() before $population / $theta / $state can be read. Calling the
# returned function directly fails with "could not find function 'ini'".
mods <- lapply(models, function(m) rxode2::rxode(readModelDb(m)))
names(mods) <- models
knitr::kable(
  data.frame(
    Model = models,
    Species = vapply(mods, function(u) u$population$species, ""),
    States = vapply(mods, function(u) length(u$state), 0L),
    Endpoints = vapply(mods, function(u) nrow(u$predDf), 0L),
    row.names = NULL
  ),
  caption = "The five packaged models. Three mouse fits (one per dietary iron status), plus the rat and human FCM extrapolations."
)
```

```{r citation}
cat(mods[[1]]$reference)
```

# Population

The mouse model was fitted to digitised group-mean PK / biodistribution data
(reference [23] of the paper). Iron-deficient, iron-adequate and iron-loaded
mice were produced by feeding diets of differing iron content, then given a
single intravenous bolus of 0.2 umol/kg `59`Fe and killed at 15 min, 12 h, 24 h
and days 4, 7, 14 and 28. Body weight was 25 g. Only means were available, so
the authors used a naive pooled-data method, treating all observations as one
individual; consequently **no between-subject variability was estimated** (all
seven `$OMEGA` elements are `0 FIX`) and the packaged models carry no etas.

The rat cohort (reference [24]) had iron-deficiency anaemia and received a
single 30 mg Fe/kg IV dose of FCM (345 g body weight). The human cohort
(reference [25]) were patients with iron-deficiency anaemia given single
ascending IV FCM doses of 100, 500, 800 and 1000 mg Fe; human physiology was
fixed to a 73 kg adult.

```{r population}
str(mods[["Fan_2025_iron_mouse_ironadequate_pbpk"]]$population)
```

# Source trace

Every `ini()` entry carries an in-file comment naming its source location. The
table below collects them.

| Equation / parameter | Value | Source location |
|---|---|---|
| Perfusion-limited organ, `d/dt(<organ>)` | n/a | Eq. (1), p. 1112 |
| Bone, minus erythropoietic utilisation | n/a | Eq. (2), p. 1112 |
| Red-cell pool, `d/dt(rbc_iron)` | n/a | Eq. (3), p. 1112 |
| Spleen, erythrophagocytic return | n/a | Eq. (4), p. 1113 |
| Liver, portal return from gut + spleen | n/a | Eq. (5), p. 1113 |
| Plasma / arterial | n/a | Eq. (6), p. 1113 |
| Residual error, proportional / additive | n/a | Eqs. (7) and (8), p. 1114 |
| `QE(human)` RBC-lifespan allometry | b = 0.75 | Eq. (9), p. 1114 |
| Full ODE system as fitted | n/a | Supplement, NONMEM `$DES` |
| `kp_*` (12 organs), mouse | Table 2 | Table 2, p. 1115 |
| `cl` (`CLloss`), mouse | Table 2 | Table 2, p. 1115 |
| `q_bone_rbc` (`QE`), mouse | Table 2 | Table 2, p. 1115 |
| `mtt_rbc` (`TRBC`), mouse | Table 2 | Table 2, p. 1115 |
| Mouse organ volumes / blood flows | Table 1 | Table 1, p. 1111 and `$PK` |
| Mouse cardiac output | `0.275 * BW^0.75` L/min | `$PK` (Brown 1988) |
| `kp_*` estimated in the rat (6 organs) | 8.64, 59.4, 147.9, 17.64, 2.28, 6.41 | Results text, p. 1116 |
| Rat organ volumes / blood flows | Supplementary Table 1 | Supplement |
| Human organ volumes / blood flows | Supplementary Table 2 | Supplement |
| `mtt_rbc` (human) | 120 days | Methods, p. 1114 |
| Rat / human body weight, cardiac output | 0.345 / 73 kg; 6.624 / 336 L/h | Methods, p. 1114 |
| FCM release limb, `KA := q_spleen` | n/a | Methods, p. 1114 (prose only) |
| Observed datasets (mouse x3, rat, human) | n/a | Supplement, "DATA FILES" |

Three `ini()` values are **not** printed in the paper and are flagged inline in
the model files; see *Assumptions and deviations* below.

# Structural verification

Before comparing against data, two structural properties of the published
`$DES` are checked directly.

```{r massbalance}
u <- mods[["Fan_2025_iron_mouse_ironadequate_pbpk"]]
p <- rxode2::rxSolve(
  u, rxode2::et(amt = 5, cmt = "plasma") |> rxode2::et(c(0, 1)),
  atol = 1e-12, rtol = 1e-10, returnType = "data.frame"
)[1, ]

# The fixed physiology lives in ini(), so it is a PARAMETER: it does not appear
# as a column in the rxSolve output. Only quantities derived inside model()
# (qc, the q_* flows, v_other) come back as columns.
th <- u$theta
fixed_vols <- c("v_bone", "v_gut", "v_heart", "v_kidney", "v_liver", "v_muscle",
                "v_skin", "v_lung", "v_spleen", "v_plasma", "v_adipose",
                "v_brain")

efflux <- with(p, q_other + q_bone + q_heart + q_kidney + q_liver +
                 q_muscle + q_skin + q_brain + q_adipose)
volsum <- sum(th[fixed_vols]) + p$v_other

knitr::kable(data.frame(
  Check = c(
    "Cardiac output (L/h)",
    "Plasma efflux coefficient, excluding cl (L/h)",
    "Efflux / cardiac output",
    "Sum of all 13 compartment volumes (L)",
    "Declared total body volume v_body (L)"
  ),
  Value = c(
    sprintf("%.5f", p$qc), sprintf("%.5f", efflux),
    sprintf("%.10f", efflux / p$qc), sprintf("%.5f", volsum),
    sprintf("%.5f", th[["v_body"]])
  )
), caption = "Plasma mass balance. The efflux coefficient of Eq. (6) equals the cardiac output exactly, and the 13 volumes reconstitute the declared body volume, confirming the rest-of-body derivation used here.")
```

The plasma compartment balances to the cardiac output **exactly**, which is the
evidence that `q_other` must be computed without subtracting gut and spleen a
second time (they are already inside `q_liver`). Table 1's printed Remainder
blood flow of 0.176 L/h does subtract them twice; see the Errata.

The lung is a different matter. Reproduced verbatim from `$DES`, it draws the
full cardiac output out of plasma and returns nothing, yet `q_lung` appears in
neither the plasma efflux sum nor the plasma influx sum:

```{r lung}
cat(sprintf(
  "q_lung = %.5f L/h, but q_lung is absent from the plasma ODE.\n%s\n",
  p$q_lung,
  "The lung is therefore a non-mass-balanced observer compartment as published."
))
```

This is kept as fitted rather than silently corrected, because the Table 2
partition coefficients were estimated against exactly this system.

# Mouse: reproducing Figures 3 and 4

The observed data are the digitised group means transcribed in the supplement's
"DATA FILES" section, in nmol of `59`Fe. Note that the supplement tabulates
times in **days** while the model's rate constants are per **hour**, so the
observation times are multiplied by 24.

```{r mouse-obs}
mouse_obs <- bind_rows(
  data.frame(diet = "iron-deficient", time = c(0.5, 1, 4, 7, 14, 28),
    plasma = c(.1088,.05576,.02312,.02312,.01632,.02312),
    bone = c(1.8954,.3978,.297,.1728,.0738,.1152),
    gut = c(.44894,.29366,.2344,.13146,.07906,.06134),
    heart = c(.0462,.0378,.0518,.0462,.0364,.0574),
    kidney = c(.1368,.1102,.1368,.057,.057,.0722),
    liver = c(.68198,.60268,.65758,.31598,.2806,.13176),
    muscle = c(.4026,.5368,.4026,.4026,.2684,.671),
    skin = c(.0124,.0186,.0155,.0186,.0124,.0093),
    lung = c(.1157,.338,.1586,.0364,.0624,.0208),
    spleen = c(.09079,.03017,.00595,NA,NA,NA),
    rbc_iron = c(.6248,1.42,1.8744,1.9312,2.0732,1.9596),
    brain = c(.0094,.0094,.0141,.0188,.0188,.0188),
    adipose = c(.0124,.0186,.0155,.0186,.0124,.0093)),
  data.frame(diet = "iron-adequate", time = c(0.5, 1, 4, 7, 14, 28),
    plasma = c(.11832,.0204,.01496,.0068,.01768,.02176),
    bone = c(3.4866,1.296,.6858,.5076,.4518,.2232),
    gut = c(.4514,.366,.1952,.061,.0488,.0366),
    heart = c(.0728,.0364,.056,.056,.0518,.0322),
    kidney = c(.19,.1178,.133,.0646,.0646,.0684),
    liver = c(1.03578,.86132,.81008,.66734,.52704,.31476),
    muscle = c(.8052,.4026,.4026,1.0736,.1342,.2684),
    skin = c(.5955,.397,.5558,.5955,.7146,.2779),
    lung = c(.1014,.1521,.0637,.0429,.0429,.039),
    spleen = c(.05817,.08071,.04333,.04074,.02786,.03066),
    rbc_iron = c(.6106,1.2283,1.5762,1.2638,1.2496,1.1644),
    brain = c(.0141,.0094,.0141,.0094,.0141,.0094),
    adipose = c(.0124,.0186,.0155,.0341,.0155,.0031)),
  data.frame(diet = "iron-loaded", time = c(0.5, 1, 4, 7, 14, 28),
    plasma = c(.1156,.03536,.00816,.00544,.00952,.01768),
    bone = c(1.9116,1.2348,.4212,.3438,.5472,.3168),
    gut = c(.28426,.47214,.20374,.10126,.09394,.07686),
    heart = c(.0938,.1134,.0896,.1036,.042,.0378),
    kidney = c(.285,.646,.38,.1786,.1558,.133),
    liver = c(.93818,2.84504,1.8361,1.0248,1.0919,1.09556),
    muscle = c(.5368,1.0736,.5368,.5368,.5368,.2684),
    skin = c(.397,.794,.3176,.397,.3573,.3573),
    lung = c(.1833,.1742,.1898,.3965,.1638,.104),
    spleen = c(.03682,.06902,.04431,.04963,.04424,.06496),
    rbc_iron = c(.1704,.3692,.568,.852,1.3064,.9088),
    brain = c(.0047,.0094,.0047,.0094,.0141,.0094),
    adipose = c(.0124,.0279,.0186,.0217,.0217,.0062))
) |>
  pivot_longer(-c(diet, time), names_to = "compartment", values_to = "obs") |>
  filter(!is.na(obs)) |>
  mutate(time_h = time * 24)

diet_model <- c(
  "iron-deficient" = "Fan_2025_iron_mouse_irondeficient_pbpk",
  "iron-adequate"  = "Fan_2025_iron_mouse_ironadequate_pbpk",
  "iron-loaded"    = "Fan_2025_iron_mouse_ironloaded_pbpk"
)

mouse_sim <- bind_rows(lapply(names(diet_model), function(d) {
  ev <- rxode2::et(amt = 5, cmt = "plasma") |> rxode2::et(seq(0, 672, by = 4))
  s <- rxode2::rxSolve(mods[[diet_model[[d]]]], ev,
                       atol = 1e-12, rtol = 1e-10, returnType = "data.frame")
  s |>
    select(time, all_of(unique(mouse_obs$compartment))) |>
    pivot_longer(-time, names_to = "compartment", values_to = "pred") |>
    mutate(diet = d)
}))
```

```{r mouse-fig, fig.height = 7.5, fig.cap = "Replicates Figures 3 and 4 of Fan 2025: observed (points, digitised group means) and model-predicted (lines) 59Fe amounts in plasma and 12 tissues after a single 0.2 umol/kg IV bolus, under three dietary iron statuses."}
ggplot(mouse_sim, aes(time / 24, pred, colour = diet)) +
  geom_line() +
  geom_point(data = mouse_obs, aes(time, obs, colour = diet), size = 1.1) +
  facet_wrap(~compartment, scales = "free_y", ncol = 4) +
  scale_x_continuous(breaks = c(0, 7, 14, 21, 28)) +
  labs(x = "Time (days)", y = "59Fe amount (nmol)", colour = "Diet") +
  theme_bw(base_size = 9) +
  theme(legend.position = "top")
```

```{r mouse-agreement}
# Interpolate per (diet, compartment) group. Deliberately NOT a left_join:
# mouse_obs and mouse_sim both carry a `time` column, so a join renames them to
# time.x / time.y and a bare `time` inside approx() would silently resolve to
# stats::time() instead of the data.
mouse_cmp <- do.call(rbind, lapply(
  split(mouse_obs, list(mouse_obs$diet, mouse_obs$compartment), drop = TRUE),
  function(o) {
    s <- mouse_sim[mouse_sim$diet == o$diet[1] &
                     mouse_sim$compartment == o$compartment[1], ]
    o$pred <- approx(s$time, s$pred, xout = o$time_h)$y
    o
  }
))
mouse_cmp$ratio <- mouse_cmp$pred / mouse_cmp$obs
stopifnot(nrow(mouse_cmp) == nrow(mouse_obs), !anyNA(mouse_cmp$pred))

knitr::kable(
  mouse_cmp |>
    group_by(Diet = diet) |>
    summarise(
      n = n(),
      `Geometric mean pred/obs` = sprintf("%.3f", exp(mean(log(ratio)))),
      `Mean |log10 ratio|` = sprintf("%.3f", mean(abs(log10(ratio)))),
      `Within 2-fold` = sprintf("%.0f%%", 100 * mean(abs(log10(ratio)) < log10(2)))
    ),
  caption = "Mouse model agreement across plasma and 12 tissues. A geometric mean pred/obs near 1 with 80-88% of points inside 2-fold is close agreement for a 13-endpoint simultaneous fit to digitised means."
)
```

All three diets reproduce the published fit: the geometric mean predicted /
observed ratio is 1.07 to 1.15 and 80-88 % of the 75-78 observations fall inside
two-fold. The model also reproduces the paper's mechanistic conclusions: the
iron-deficient fit has the highest `q_bone_rbc` and the lowest `cl`, driving
`59`Fe into the red-cell pool, while the iron-loaded fit has the largest liver,
heart and kidney partition coefficients.

```{r mouse-params}
knitr::kable(
  data.frame(
    Diet = names(diet_model),
    `q_bone_rbc (L/h)` = sprintf("%.2e", vapply(diet_model, function(m)
      exp(mods[[m]]$theta[["lq_bone_rbc"]]), 0)),
    `cl (L/h)` = sprintf("%.3e", vapply(diet_model, function(m)
      exp(mods[[m]]$theta[["lcl"]]), 0)),
    `kp_liver` = sprintf("%.2f", vapply(diet_model, function(m)
      exp(mods[[m]]$theta[["lkp_liver"]]), 0)),
    `mtt_rbc (h)` = sprintf("%.2f", vapply(diet_model, function(m)
      exp(mods[[m]]$theta[["lmtt_rbc"]]), 0)),
    row.names = NULL, check.names = FALSE
  ),
  caption = "Table 2 parameters that carry the paper's mechanistic argument, read back from the packaged models."
)
```

# Rat: reproducing Figure 5

The rat and human models add one circulating `ivip` compartment: the whole IV
FCM dose lands there and is taken up by splenic macrophages at a rate set equal
to the spleen blood flow (`KA := q_spleen`). Observations are on total serum
iron, `Cc = (plasma + ivip) / v_plasma`.

Both extrapolation models declare several endpoints, so observation records must
identify which endpoint they belong to. `dvid = 1` selects the first endpoint
(`Cc`); this is the `dvid()` route rather than writing `cmt = "Cc"`, which would
reference an algebraic observable as if it were a compartment.

```{r ev-helper}
ev_ivip <- function(dose, times) {
  # Keep t = 0 in the observation grid. The dose record comes first, so the
  # t = 0 observation reports the POST-dose state and Cc(0) = dose / v_plasma
  # exactly -- which is the dose-to-serum mapping check used below. Dropping
  # t = 0 silently loses that row and makes the mapping table empty.
  times <- sort(unique(c(0, times[times >= 0])))
  data.frame(
    id   = 1L,
    time = c(0, times),
    amt  = c(dose, rep(NA_real_, length(times))),
    evid = c(1L, rep(0L, length(times))),
    cmt  = c("ivip", rep(NA_character_, length(times))),
    dvid = c(NA_integer_, rep(1L, length(times)))
  )
}
```

```{r rat}
rat_serum <- data.frame(
  time = c(1, 2, 4, 6, 8, 12, 24, 48),
  obs = c(1147.87, 979.73, 625.86, 839.33, 660.93, 526.43, 438.72, 111.29)
)
rat_tissue <- data.frame(time = c(72, 192, 360),
  Cbone = c(42.40, 38.99, 15.27), Cheart = c(64.54, 51.70, 40.42),
  Ckidney = c(36.65, 25.70, 19.92), Cliver = c(341.36, 236.92, 84.11),
  Cmuscle = c(11.83, 11.27, 7.70), Cspleen = c(449.76, 461.72, 385.17)
) |> pivot_longer(-time, names_to = "compartment", values_to = "obs")

ur <- mods[["Fan_2025_iron_ferriccarboxymaltose_rat_pbpk"]]
sr <- rxode2::rxSolve(
  ur, ev_ivip(10.35, c(seq(0, 400, by = 0.25), rat_serum$time, rat_tissue$time)),
  atol = 1e-12, rtol = 1e-10, returnType = "data.frame"
)

rat_cmp <- bind_rows(
  rat_serum |>
    mutate(compartment = "serum (ug/dL)",
           pred = approx(sr$time, sr$Cc * 100, xout = time)$y),
  rat_tissue |>
    group_by(compartment) |>
    mutate(pred = approx(sr$time, sr[[first(compartment)]], xout = time)$y) |>
    ungroup()
) |> mutate(ratio = pred / obs)

knitr::kable(
  rat_cmp |>
    mutate(Group = if_else(compartment == "serum (ug/dL)", "Serum", "Tissue")) |>
    group_by(Group) |>
    summarise(n = n(),
      `Geometric mean pred/obs` = sprintf("%.2f", exp(mean(log(ratio)))),
      `Mean |log10 ratio|` = sprintf("%.2f", mean(abs(log10(ratio)))),
      `Within 2-fold` = sprintf("%.0f%%", 100 * mean(abs(log10(ratio)) < log10(2)))),
  caption = "Rat extrapolation agreement (Figure 5). Serum is 30 mg Fe/kg FCM; tissues are bone, heart, kidney, liver, muscle and spleen at 72, 192 and 360 h. Read the serum row together with the shape check below -- a good average ratio here does NOT mean the serum profile is reproduced."
)
```

A level-agreement statistic is not enough for the serum profile, so the *shape*
is checked separately: how far does each series fall between the first and last
serum observation?

```{r rat-shape}
rs <- filter(rat_cmp, compartment == "serum (ug/dL)")
knitr::kable(data.frame(
  Series = c("Observed (digitised)", "Predicted"),
  `First (1 h)` = sprintf("%.0f", c(rs$obs[1], rs$pred[1])),
  `Last (48 h)` = sprintf("%.0f", c(rs$obs[nrow(rs)], rs$pred[nrow(rs)])),
  `Fold decline, 1-48 h` = sprintf("%.1f-fold",
    c(rs$obs[1] / rs$obs[nrow(rs)], rs$pred[1] / rs$pred[nrow(rs)])),
  check.names = FALSE
), caption = "Rat serum shape check. The observed profile falls about 10-fold over 1-48 h; the model falls about 1.5-fold. The serum decline is NOT reproduced, and the favourable geometric mean above is a consequence of a near-flat prediction sitting in the middle of a declining observation series.")
```

```{r rat-fig, fig.height = 3.2, fig.cap = "Replicates Figure 5 of Fan 2025: observed (points) and predicted (line) serum iron in iron-deficient rats after a single 30 mg Fe/kg IV dose of ferric carboxymaltose."}
ggplot(data.frame(time = sr$time, pred = sr$Cc * 100), aes(time, pred)) +
  geom_line() +
  geom_point(data = rat_serum, aes(time, obs), size = 1.6) +
  coord_cartesian(xlim = c(0, 60)) +
  labs(x = "Time (h)", y = "Serum iron (ug/dL)") +
  theme_bw(base_size = 10)
```

**The rat tissue distribution is reproduced; the rat serum profile is not.**

The six measured tissues sit at a geometric mean predicted / observed ratio of
1.42 with 89 % of points inside two-fold - good agreement, and the substantive
result, since tissue burden is what the model exists to predict. The
over-prediction is concentrated at the last time point (360 h), consistent with
the authors' own caution about sparse terminal data.

Serum is a different story. The average level is close (geometric mean 1.10),
but that number is misleading: `KA := q_spleen` clears the FCM colloid out of rat
serum with a half-life of `ln(2) * 0.0078 / 0.0417` = 0.13 h, so more than 99 %
of the dose has left the `ivip` compartment by the first observation at 1 h.
From then on serum is set by return flux from the tissues, which is nearly flat.
The prediction therefore falls about 1.5-fold over 1-48 h against an observed
10-fold decline, and the good average ratio is an artifact of a flat line
crossing a declining one. Unlike the human case below, refitting `q_spleen`
does **not** repair the rat profile (the implied value is within a few percent
of the published one), because the rat serum level is governed by tissue return
rather than by colloid release. The authors report only that *"the iron
concentration in the serum was slightly overestimated to a lesser degree after
20 h"*; the discrepancy is in the shape, not the level.

# Human: reproducing Figure 6

```{r human}
human_obs <- data.frame(
  time = c(.25,.5,.75,1,1.5,2,3,4,6,8,12,16,24,36,48,60,72),
  `100` = c(36.3,36.1,34.73,33.17,31.69,31.27,26.97,25.50,20.92,17.97,
            13.66,12.59,8.82,7.75,7,NA,NA),
  `500` = c(153.23,150,148.39,141.94,140.32,133.87,127.42,122.58,101.61,88.71,
            69.35,51.61,30.65,12.90,8.06,7,NA),
  `800` = c(287.72,292.61,272.89,287.63,292.47,262.90,285.71,239.67,218.10,
            185.03,141.87,124.93,79.59,29.87,11.57,8.74,7),
  `1000` = c(322.87,NA,308.37,318.46,320.78,308.09,297.72,288.99,261.61,240.79,
             215.71,175.79,135.21,78.74,54.64,22.10,7),
  check.names = FALSE
) |>
  pivot_longer(-time, names_to = "dose", values_to = "obs") |>
  filter(!is.na(obs)) |>
  mutate(dose_mg = as.numeric(dose))

uh <- mods[["Fan_2025_iron_ferriccarboxymaltose_human_pbpk"]]
human_times <- c(human_obs$time, seq(0, 80, by = 0.1))
human_sim <- bind_rows(lapply(c(100, 500, 800, 1000), function(d) {
  s <- rxode2::rxSolve(uh, ev_ivip(d, human_times), atol = 1e-14, rtol = 1e-12,
                       returnType = "data.frame")
  data.frame(time = s$time, pred = s$Cc, dose = as.character(d), dose_mg = d)
}))
```

```{r human-fig, fig.height = 3.6, fig.cap = "Replicates Figure 6 of Fan 2025: observed (points, digitised) and predicted (lines) total serum iron after single ascending IV doses of ferric carboxymaltose. The predicted peak matches, but the predicted decline is far too fast -- see the Errata."}
ggplot(human_sim, aes(time, pred, colour = dose)) +
  geom_line() +
  geom_point(data = human_obs, aes(time, obs, colour = dose), size = 1.4) +
  scale_y_log10() +
  coord_cartesian(xlim = c(0, 72), ylim = c(1, 500)) +
  labs(x = "Time (h)", y = "Total serum iron (ug/mL)", colour = "Dose (mg)") +
  theme_bw(base_size = 10)
```

Two distinct checks are needed here, and conflating them is easy. The first asks
whether the *dose-to-serum mapping* is right: does the whole dose show up in the
serum volume immediately after the bolus? The second asks whether the model
matches the data *at the times the data were actually collected*, the first of
which is 0.25 h.

```{r human-mapping}
h0 <- human_sim |> filter(time == 0) |> select(dose_mg, cc0 = pred)
knitr::kable(
  h0 |> transmute(`Dose (mg)` = dose_mg,
                  `dose / v_plasma = dose / 3 L (ug/mL)` = sprintf("%.1f", dose_mg / 3),
                  `Model Cc at t = 0+ (ug/mL)` = sprintf("%.1f", cc0),
                  `Ratio` = sprintf("%.3f", cc0 / (dose_mg / 3))),
  caption = "Dose-to-serum mapping check. Immediately after the bolus the model holds the entire dose in the 3 L serum volume, so Cc(0+) = dose / 3 exactly. This confirms the dosing route and the observation definition, and it is the right order of magnitude for the observed peaks (36-323 ug/mL) -- but t = 0+ is not an observable sample."
)
```

```{r human-matched}
human_matched <- human_obs |>
  group_by(dose_mg) |>
  group_modify(function(.x, .y) {
    s <- human_sim[human_sim$dose_mg == .y$dose_mg, ]
    .x$pred <- approx(s$time, s$pred, xout = .x$time)$y
    .x
  }) |>
  ungroup() |>
  mutate(ratio = pred / obs)

knitr::kable(
  human_matched |>
    group_by(`Dose (mg)` = dose_mg) |>
    summarise(
      n = n(),
      `Obs at 0.25 h` = sprintf("%.1f", obs[which.min(time)]),
      `Pred at 0.25 h` = sprintf("%.1f", pred[which.min(time)]),
      `Ratio at 0.25 h` = sprintf("%.2f", (pred / obs)[which.min(time)]),
      `Geomean pred/obs, all times` = sprintf("%.3f", exp(mean(log(ratio)))),
      .groups = "drop"
    ),
  caption = "Human serum at MATCHED observation times. Even the earliest sample at 0.25 h is already about 2.2-fold low, and averaged over the whole profile the model is roughly 30-fold low. The peak is NOT reproduced at any sampled time."
)
```

So the dose-to-serum mapping is right, but **the profile is not reproduced at
any observed time**, including the first. The reason is that splenic macrophage
uptake is set to the human spleen blood flow, `q_spleen` = 10.08 L/h, giving an
intact-colloid serum half-life of `ln(2) * 3 / 10.08` = 0.21 h; by the 0.25 h
sample the model has already shed more than half the dose out of serum, and by
1 h it has shed 96 % of it. The observed profiles instead decline with a
terminal half-life of roughly 10-15 h. Fitting `q_spleen` as a free parameter to
the 1000 mg profile - a **diagnostic only**, not a change to the packaged
model - recovers the data closely:

```{r human-ka}
h1000 <- filter(human_obs, dose_mg == 1000)
ev1000 <- ev_ivip(1000, c(h1000$time, seq(0, 80, by = 0.25)))
obj <- function(lka) {
  s <- rxode2::rxSolve(uh, c(q_spleen = exp(lka)), ev1000,
                       atol = 1e-12, rtol = 1e-10, returnType = "data.frame")
  sum((log(pmax(approx(s$time, s$Cc, xout = h1000$time)$y, 1e-9)) -
         log(h1000$obs))^2)
}
o <- optimize(obj, c(log(1e-3), log(50)))
sfit <- rxode2::rxSolve(uh, c(q_spleen = exp(o$minimum)), ev1000,
                        atol = 1e-12, rtol = 1e-10, returnType = "data.frame")
rfit <- approx(sfit$time, sfit$Cc, xout = h1000$time)$y / h1000$obs
knitr::kable(data.frame(
  Quantity = c("Published KA := q_spleen (L/h)",
               "Implied effective KA from the 1000 mg profile (L/h)",
               "Ratio", "Geometric mean pred/obs with the implied KA",
               "Mean |log10 ratio| with the implied KA"),
  Value = c("10.08", sprintf("%.3f", exp(o$minimum)),
            sprintf("%.0f-fold slower", 10.08 / exp(o$minimum)),
            sprintf("%.2f", exp(mean(log(rfit)))),
            sprintf("%.2f", mean(abs(log10(rfit)))))
), caption = "Diagnostic only. The structure and the dose-to-serum mapping are correct; the single simplification KA := q_spleen is about 72-fold too fast in humans. The packaged model retains the published value.")
```

So the human file is a faithful encoding of the published recipe, and it is
internally consistent and exactly dose-proportional, but it does **not**
reproduce the observed serum time course at any sampled time. This is a property
of the paper, not of the extraction; it is recorded in the Errata below.

# NCA validation (PKNCA)

The paper reports no NCA table, so the simulated human profiles are compared
against NCA of the **digitised observed** profiles. Both series are put on the
**same time grid** - the actual observation times - and the interval starts at
the first measurement (0.25 h), so neither series needs a fabricated time-zero
record and PKNCA raises no "AUC range starting before the first measurement"
warning. This makes the comparison a direct statement about the profiles rather
than about an extrapolation choice.

`half.life` is deliberately not requested: the simulated profile is
non-monotonic (a fast collapse followed by a slow rebound plateau fed by tissue
return), so a terminal slope is not a meaningful quantity for it.

```{r pknca}
nca_frame <- function(df, value) {
  df |>
    transmute(id = 1L, treatment = paste(dose_mg, "mg"), time,
              Cc = .data[[value]]) |>
    filter(!is.na(Cc)) |>
    arrange(treatment, time)
}

run_nca <- function(conc_df, doses) {
  conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | treatment + id,
                               concu = "ug/mL", timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(doses, amt ~ time | treatment + id,
                               doseu = "mg")
  # tmax is omitted: both series peak at (or one sample after) the first
  # record of an IV bolus profile that starts at 0.25 h, so it carries no
  # information here and PKNCA reports it relative to the interval start.
  intervals <- data.frame(start = min(conc_df$time), end = 48,
                          cmax = TRUE, auclast = TRUE)
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}

dose_df <- data.frame(id = 1L, time = 0,
                      amt = c(100, 500, 800, 1000),
                      treatment = paste(c(100, 500, 800, 1000), "mg"))

nca_obs <- run_nca(nca_frame(filter(human_matched, time <= 48), "obs"), dose_df)
nca_sim <- run_nca(nca_frame(filter(human_matched, time <= 48), "pred"), dose_df)
```

```{r pknca-table}
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_sim,
  reference = nca_obs,
  by = "treatment",
  params = c("cmax", "auclast"),
  units = c(cmax = "ug/mL", auclast = "ug*h/mL"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = "Simulated vs digitised-observed NCA of human total serum iron on matched observation times, 0.25-48 h. Every Cmax and AUClast row is starred: the packaged model is roughly 2-fold low at the earliest sample and roughly 25-fold low on exposure. Starred rows differ by more than 20%."
)
```

**Every Cmax and AUClast row is starred.** This is the expected, documented
consequence of `KA := q_spleen` (see the Errata), not a signal to tune any
parameter. It is recorded here rather than suppressed, because a reader deciding
whether to use the human file for an exposure calculation needs to see it.

What the model *does* get right is strict dose proportionality, which is worth
confirming separately since it is the property the file remains usable for:

```{r dose-prop}
# Integrate on the COMMON simulation grid, not on the per-dose observed times:
# the four observed series have different sampling schedules and different last
# observed times, so a trapezoid over them is not comparable across doses and
# would not test linearity at all.
dose_prop <- human_sim |>
  filter(time <= 48) |>
  arrange(dose_mg, time) |>
  group_by(dose_mg) |>
  summarise(
    auc = sum(diff(time) * (head(pred, -1) + tail(pred, -1)) / 2),
    .groups = "drop"
  ) |>
  mutate(auc_per_dose = auc / dose_mg)

knitr::kable(
  dose_prop |>
    transmute(`Dose (mg)` = dose_mg,
              `Simulated AUC 0-48 h (ug*h/mL)` = sprintf("%.2f", auc),
              `AUC / dose` = sprintf("%.6f", auc_per_dose)),
  caption = "Simulated exposure on a common 0-48 h grid is exactly proportional to dose (constant AUC / dose), as expected for a linear system dosed into a single serum volume. The model is internally consistent; it is the absolute level of the serum time course that does not match the data."
)
stopifnot(diff(range(dose_prop$auc_per_dose)) / mean(dose_prop$auc_per_dose) < 1e-6)
```

# Assumptions and deviations

Three values in the rat and human files are not printed anywhere in the paper
or its supplement. Each is flagged inline in the model file as well as here.

1. **Which mouse iron-status column the extrapolations inherit from.** The paper
   says only that the non-estimated rat and human partition coefficients and
   `CLloss` "were assumed to be identical ... in mice", and Eq. (9) needs a
   `QE(mouse)` and a `TRBC(mouse)`. Table 2 offers three columns and the paper
   never says which was used. The **iron-adequate** column is used here, as the
   reference physiological state conventionally used for cross-species scaling
   (operator ruling, sidecar q1 = B). This choice propagates into *both* the rat
   and the human file: `QE(human)` would be 1.04e-2 L/h from the iron-deficient
   column instead of the 6.00e-3 L/h used here, a 1.7-fold difference, and
   `kp_liver` for the unmeasured-tissue inheritance would be 9.38 rather than
   18.11. Rat and human predictions would therefore shift if the iron-deficient
   column was in fact the authors' basis - which is arguable, since both
   validation cohorts had iron-deficiency anaemia.
2. **The rat's `mtt_rbc` and `q_bone_rbc`.** Neither is reported for the rat
   anywhere on disk, yet both are load-bearing for the bone and spleen profiles
   of Figure 5. Per the operator ruling (sidecar q2 = B) they were obtained by
   applying the paper's **own** Eq. (9) allometry rather than by copying the
   printed mouse values. The rat red-cell lifespan is derived from the paper's
   two lifespan anchors - mouse iron-adequate `TRBC` = 34.44 h at 0.025 kg and
   human `TRBC` = 120 days at 73 kg - which fix an exponent
   `b_T = log(2880 / 34.44) / log(73 / 0.025)` = 0.5547, evaluated at the
   paper's rat weight of 0.345 kg to give 147.7 h; `q_bone_rbc` then follows
   from Eq. (9) with b = 0.75, giving 6.47e-4 L/h. The derivation uses only
   published numbers, but it is a derivation, not a published rat value.
3. **The FCM / IVIP release topology is a structural assumption.** The paper
   describes the limb only in prose - "the FCM absorption rate KA was
   incorporated as the spleen blood flow rate" - and the control stream in the
   supplement is the mouse, pure-iron run, so it contains no such limb. Per the
   operator ruling (sidecar q3 = A) the released iron is routed into the
   **spleen** compartment, on the physiological grounds that iron-carbohydrate
   colloids are cleared by reticuloendothelial macrophages, of which the spleen
   is the principal accessible site in this structure. Routing it straight to
   plasma instead would bypass the macrophage step the limb exists to represent.
   The paper does not state which is intended.

Beyond those three, one finding about the paper matters more than any of them
for a user deciding what these files are good for.

**`KA := q_spleen` is far too fast, so neither extrapolation model reproduces
its serum profile.** Setting the FCM colloid uptake rate equal to the spleen
blood flow gives a serum half-life for the intact colloid of
`ln(2) * v_plasma / q_spleen` - 0.13 h in the rat and 0.21 h in the human. The
observed serum profiles evolve over tens of hours. The consequence differs by
species, and the two diagnoses are distinct:

- **Human.** The dose-to-serum *mapping* is right - immediately after the bolus
  the model holds the whole dose in 3 L, giving `Cc(0+) = dose / 3`, the right
  order of magnitude for the observed 36-323 ug/mL peaks - but that value is at
  `t = 0+` and is not an observable sample. At the **matched** observation times
  the model is already about 2.2-fold low at the earliest sample (0.25 h) and
  about 30-fold low averaged over the profile, so neither Cmax nor AUC is
  reproduced. This one *is* a `KA` problem: refitting `q_spleen` alone against
  the 1000 mg profile gives 0.139 L/h - about 72-fold slower - and recovers the
  data to a geometric mean predicted / observed of 0.96 (mean absolute log10
  ratio 0.05).
- **Rat.** The tissue distribution is right, but serum settles onto a near-flat
  plateau set by return flux from the tissues, falling about 1.5-fold over
  1-48 h against an observed 10-fold decline. Refitting `q_spleen` does **not**
  help here (the implied value is within a few percent of the published one), so
  the rat serum discrepancy has a different cause - the model has no sink able
  to produce a 10-fold serum decline on that timescale.

Both refits above are **diagnostics only**; the packaged files keep the
published `KA := q_spleen`. Practical guidance: the three mouse files and the
rat *tissue* predictions are validated and can be used as they stand. The
extrapolation models' **serum** predictions are not: do not use either for serum
Cmax, AUC or half-life. The human file remains a faithful, internally consistent
and exactly dose-proportional encoding of the published recipe, which is what
makes it useful as a starting point for anyone who wants to re-estimate `KA`.

Further deviations and paper-side errors, all reproduced or corrected as noted:

- **Model time base is hours, not days.** The supplement tabulates the mouse
  observation times as days (0.5, 1, 4, 7, 14, 28) while every rate constant is
  per hour. Reading the times as days degrades the mouse geometric mean
  predicted / observed from 1.07 to 1.48 and under-predicts the red-cell pool
  25-fold at early times; reading them as hours after multiplying by 24
  reproduces the published fit. The packaged models therefore declare
  `units$time = "h"`. A consequence is that the fitted mouse `TRBC` of 34-42 h
  is not a biologically plausible murine red-cell lifespan (about 40 days); it
  is an empirical parameter of this fit.
- **Remainder blood flow, all three physiology tables.** Table 1 (mouse,
  0.176 L/h), Supplementary Table 1 (rat, 3.584 L/h) and Supplementary Table 2
  (human, 28.69 L/h) all subtract gut and spleen flow a second time, although
  both are already inside the total liver flow. The fitted `$PK` code does not:
  it uses `QROB = QCO - QBON - QHEA - QKID - QLIV - QMUS - QSKI - QADI - QBRA`,
  which is what makes the plasma compartment balance exactly to cardiac output
  (verified above). The packaged models follow the code, giving 0.333, 4.125 and
  89.17 L/h. The systematic recurrence across all three tables identifies this
  as a table-generation artifact rather than three typos.
- **Lung mass balance.** Reproduced as published: the lung draws the full
  cardiac output from plasma and returns nothing, but `q_lung` is in neither the
  plasma efflux nor the plasma influx sum of `$DES`. It is a non-mass-balanced
  observer compartment. Kept as fitted, because Table 2 was estimated against
  this system.
- **Cardiac output, mouse.** Methods state "The Q for the mouse applied in the
  model was 0.72 L/h", but `$PK` computes `60 * 0.275 * 0.025^0.75` = 1.037 L/h,
  which is exactly the Table 1 lung flow of 1.04 and exactly the sum of the
  tabulated tissue flows. The 1.037 L/h value is the one that was fitted.
- **Mouse dose.** Figure 3's caption says 0.02 umol/kg but Methods says
  0.2 umol/kg. The supplement's data tables start plasma at 5 nmol, and
  0.2 umol/kg x 0.025 kg = 5 nmol, confirming Methods.
- **Residual error values are initial estimates.** The `$ERROR` block fully
  determines the *structure* - proportional-only for plasma, bone, gut, skin,
  lung, spleen and brain; additive-only for heart, kidney, liver, muscle,
  red-cell iron and fat - but the paper reports no final residual estimates
  anywhere, so the packaged values are the `$SIGMA` **initial** entries
  (variances 0.1, and 0.5 for bone), converted to nlmixr2 standard deviations.
  For the rat and human files, which have no control stream at all, the mouse
  proportional initial is carried forward; the mouse additive arms are on a nmol
  amount scale and are not transferable to the mg scale of those models, so
  proportional error is used throughout there.
- **Observations are amounts in the mouse, concentrations in the rat and
  human.** The published mouse `$ERROR` sets `IPRED = A(n)`, so the mouse
  endpoints are amounts in nmol and the red-cell pool has no volume at all. The
  rat tissue data are reported in mg/kg and the rat and human serum data in
  concentration units, so those endpoints are concentrations, assuming unit
  tissue density.
- **No between-subject variability.** All seven `$OMEGA` elements are `0 FIX`
  (naive pooled fit to group means), so no etas are encoded. This is a property
  of the source, not an omission.
- **Stiffness.** With `kp_other` estimated at 2.3e-9 in the iron-adequate fit,
  the rest-of-body efflux rate constant is of order 1e10 per hour. All
  simulations here set `atol` and `rtol` tightly; loose tolerances will produce
  visible integration noise.
- **New canonical names.** `q_bone_rbc` (the paper's `QE`) and `mtt_rbc` (the
  paper's `TRBC`) are registered in `inst/references/parameter-names.md`, as a
  member of the existing `q_<from>_<to>` one-way inter-compartmental clearance
  family and of the `mtt_<celltype>` cell-lifespan family respectively. The
  paper's `CLloss` is encoded as the canonical `cl`, since it is the model's
  only elimination pathway. The circulating FCM colloid is declared as a
  `paper_specific_compartments` entry (`ivip`) rather than as a new canonical
  compartment.

# Session information

```{r session}
sessionInfo()
```
