---
title: "Pramlintide (Fang 2013)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pramlintide (Fang 2013)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Fang_2013_pramlintide"))$reference`
* Description: `r rxode2::rxode(readModelDb("Fang_2013_pramlintide"))$description`
* Article: <https://doi.org/10.1208/s12248-012-9409-7>
* Open-access full text: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3535104/>

Fang 2013 is a mechanism-based reanalysis of the Colburn 1996 AC137 crossover
study. It is the first published mechanism-based PK/PD model for pramlintide,
an amylinomimetic used as an adjunct to insulin in type 1 diabetes. The paper
publishes both NONMEM control streams in its Appendix (pages 25 to 28), so the
structural model below is transcribed from the control streams and the printed
equations together, with parameter values from Tables II and III.

# Population

Twenty-five men with type 1 diabetes mellitus took part in a randomised,
single-blind, placebo-controlled ascending-dose crossover trial. Mean (SD) age
was 29.6 (6.8) years, body weight 170.4 (17.0) lb (77.3 (7.7) kg) and height
71.6 (2.5) in. All subjects were white except one Hispanic subject. Diabetes had
been diagnosed 2 to 20 years earlier, glycosylated haemoglobin ranged from 6.1
to 13.0 percent and basal C-peptide was below 0.6 ng/mL.

Subjects were allocated to one of three dose groups (30, 100 or 300 ug
pramlintide). Within a dose group each subject received the drug twice, once as
a nominal 2 min intravenous bolus and once as a 120 min intravenous infusion,
in a two-period crossover, and each period embedded a further placebo-versus-
active crossover on consecutive days. Two subjects per dose group received
placebo throughout. Subjects took their usual pre-breakfast insulin about 30 min
before dose initiation and breakfast was served 30 min after dose initiation
(Fang 2013 Methods, Study Design).

The pharmacokinetic model was built on 342 pramlintide observations in 18
individuals; placebo-only subjects were excluded from the PK fit because
endogenous amylin was negligible or below the limit of detection. The
pharmacodynamic model used 1028 plasma glucose observations in all 25
individuals, placebo and active occasions together.

The same information is available programmatically via
`readModelDb("Fang_2013_pramlintide")()$population`.

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Fang_2013_pramlintide.R` carries an in-file comment
pointing at its source location. They are collected here for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL) | 0.955 L/min | Table II, Structural |
| `lvc` (V1) | 19.1 L | Table II, Structural |
| `lvp` (V2) | 11.0 L | Table II, Structural |
| `lq` (CLD) | 0.283 L/min | Table II, Structural |
| `ld1` (D1) | 4 min, fixed | Table II; PK control stream `$THETA (4, FIX)` (p. 25) |
| `ltlag` (Tlag) | 0.430 min | Table II, Structural |
| `propSd` | 26.2 percent | Table II, Residual variability |
| IIV CL / V1 / V2 / D1 | 19.5 / 23.3 / 61.8 / 37.2 percent | Table II, IIV column |
| Corr(CL, V1) | 0.487 | Table II, `Corr CL_V1`; PK control stream `$OMEGA BLOCK(2)` |
| IOV CL | 10.8 percent | Table II, IOV column; `$OMEGA BLOCK(1) ... SAME` |
| `ltin` (Tin) | 85.9 min | Table III, Structural |
| `lkout` (kout) | 0.0146 1/min | Table III, Structural |
| `lka` (ka) | 0.0282 1/min | Table III, Structural |
| `lfglucose` (F) | 0.843 | Table III, Structural |
| `lrbase` (G0) | 161 mg/dL | Table III, Structural |
| `limax` (Imax) | 0.995 | Table III, Structural |
| `lic50` (IC50) | 23.8 pmol/L | Table III, Structural |
| `lsmax` (Smax) | 1.26 | Table III, Structural |
| `lvgc` (VGc) | 9.33 L, fixed | Table III footnote a (Silber 2007, reference 20) |
| `lvgp` (VGp) | 8.56 L, fixed | Table III footnote a (Silber 2007, reference 20) |
| `lqg` (QG) | 0.442 L/min, fixed | Table III footnote a (Silber 2007, reference 20) |
| `expSd_Gc` | 0.157 (log scale) | Table III, Residual error 15.7; PD control stream `W = THETA(9)*EXP(ETA(13))`, `Y = LOG(IPRED) + W*EPS(1)`, `$SIGMA 1 FIX` |
| IIV Tin / kout / F / G0 / residual | 39.7 / 83.1 / 48.5 / 38.4 / 39.5 percent | Table III, IIV column |
| IOV G0 | 37.1 percent | Table III, IOV column; `$OMEGA BLOCK(1) ... SAME` x3 |
| Meal glucose dose D | 50 g | Equation 5 and Methods, Pharmacodynamic Model |
| `d/dt(central)`, `d/dt(peripheral1)` | n/a | PK control stream `$SUBROUTINE ADVAN3`; PD control stream `$DES` `DADT(1)`, `DADT(2)` (p. 27) |
| `d/dt(intestine)` | n/a | Equation 3; `$DES` `DADT(3)` (p. 28) |
| `d/dt(glucose)` | n/a | Equation 6; `$DES` `DADT(4)` (p. 28) |
| `d/dt(glucose_per)` | n/a | Equation 7; `$DES` `DADT(5)` (p. 28) |
| `s <- 1 + smax * ON_TREATMENT` | n/a | Equation 4; PD control stream `ST1/ST2/ST3 = 1 + SMAX` (p. 26) |
| `inh <- imax * Cc / (ic50 + Cc)` | n/a | Equation 8; `$DES` `INH = IMAX*C1/(IC50+C1)` (p. 28) |
| `kin <- kout * gc0` | n/a | Equation 9; `KIN = BGLC*93.3*KOUT` (p. 27) |
| 60 min gate on the inhibition | n/a | `$DES` `IF (T.LT.60) THEN N=0 ELSE N=1` and `KIN*(1.0-N*INH)` (pp. 27 and 28) |
| `glucose(0) <- gc0` | n/a | Equation 6 initial condition; `A_0(4) = BGLC*93.3` (p. 27) |
| `glucose_per(0) <- 0.917 * rbase * vgp_dl` | n/a | `A_0(5) = 0.917*BGLC*85.6` (p. 27) |
| `dur(intestine) <- tin * s`, `f(intestine) <- fglucose` | n/a | Equations 3 and 5; PD control stream `D3`, `F3` (pp. 26 and 27) |

# Units and the dose conversion

The model carries pramlintide amounts in pmol and glucose amounts in mg, with
time in minutes. Fang 2013 reports doses in ug and concentrations in pmol/L but
does not print a molecular weight, so the ug-to-pmol conversion below uses the
published pramlintide molecular weight of 3949.4 g/mol, which is a chemical
constant and **not** a value taken from the paper. The chunk cross-checks that
constant against the paper's own Table I, where dose divided by
(AUC0-inf x CL) recovers a molecular weight independently.

```{r units}
MW_PRAMLINTIDE <- 3949.4  # g/mol; chemical constant, not from Fang 2013

ug_to_pmol <- function(ug) ug * 1e-6 / MW_PRAMLINTIDE * 1e12

# Fang 2013 Table I: dose (ug) implied by AUC0-inf (pmol*min/L) x CL (L/min).
mw_check <- tibble::tribble(
  ~arm,             ~dose_ug, ~auc,   ~cl,
  "100 ug bolus",    100,     28200,  0.923,
  "300 ug bolus",    300,     78900,  0.984,
  "100 ug infusion", 100,     25400,  1.04,
  "300 ug infusion", 300,     94100,  0.819
) |>
  dplyr::mutate(mw_implied = dose_ug * 1e-6 / (auc * cl * 1e-12))

knitr::kable(
  mw_check |>
    dplyr::rename(
      "Arm" = arm, "Dose (ug)" = dose_ug,
      "AUC0-inf (pmol*min/L)" = auc, "CL (L/min)" = cl,
      "Implied MW (g/mol)" = mw_implied
    ),
  digits = c(0, 0, 0, 3, 0),
  caption = "Molecular weight implied by Fang 2013 Table I, per dose arm."
)

# The 30 ug arms are excluded above: the paper states their profiles reached the
# limit of detection, which biases both AUC and CL for that group.
stopifnot(abs(median(mw_check$mw_implied) - MW_PRAMLINTIDE) / MW_PRAMLINTIDE < 0.10)
```

# Closed-form checks on the packaged parameters

These do not depend on a simulated cohort: they evaluate the model's own
`ini()` values against numbers Fang 2013 states in prose.

```{r closed-form}
ui <- rxode2::rxode(readModelDb("Fang_2013_pramlintide"))
th <- ui$theta

tin   <- exp(th[["ltin"]])
smax  <- exp(th[["lsmax"]])
fglc  <- exp(th[["lfglucose"]])
kout  <- exp(th[["lkout"]])
rbase <- exp(th[["lrbase"]])
vgc   <- exp(th[["lvgc"]])
cl    <- exp(th[["lcl"]])
vc    <- exp(th[["lvc"]])
vp    <- exp(th[["lvp"]])
q     <- exp(th[["lq"]])

D_MEAL_MG <- 50000  # Equation 5: D = 50 g

# Equation 5 with S = 1 (placebo). Fang 2013 Discussion: "the estimated glucose
# emptying rate is similar at about 491 mg/min".
k0_placebo <- D_MEAL_MG * fglc / tin

# Equation 4. Fang 2013 Discussion: "The estimated stimulatory effect of
# pramlintide on Tin was about 2-3-fold in this study".
s_active <- 1 + smax

# Equation 9, with Gc0 = G0 * VGc expressed in dL.
kin <- kout * rbase * vgc * 10

# Terminal disposition rate constant of the two-compartment PK model.
k10 <- cl / vc; k12 <- q / vc; k21 <- q / vp
sum_k <- k10 + k12 + k21
beta_pk <- 0.5 * (sum_k - sqrt(sum_k^2 - 4 * k10 * k21))
thalf_pk <- log(2) / beta_pk

closed <- tibble::tibble(
  Quantity = c(
    "Meal glucose emptying rate k0 under placebo (mg/min)",
    "Gastric-emptying prolongation factor S under pramlintide",
    "Endogenous glucose production kin (mg/min)",
    "Terminal half-life of pramlintide (min)"
  ),
  Value = c(k0_placebo, s_active, kin, thalf_pk),
  `Fang 2013 statement` = c(
    "about 491 mg/min (Discussion)",
    "about 2-3-fold (Discussion)",
    "not stated; equals kout x Gc0 (Equation 9)",
    "Table I reports 19.6 to 55.1 min across the six arms"
  )
)
knitr::kable(closed, digits = 3, caption = "Closed-form checks.")

stopifnot(
  abs(k0_placebo - 491) < 1,       # paper rounds to 491 mg/min
  s_active > 2 && s_active < 3,    # paper: about 2-3-fold
  thalf_pk > 19.6 && thalf_pk < 55.1
)
```

# Virtual cohort

The original observed data are not public. The cohort below reproduces the
trial's factorial design: three dose levels x two regimens x (pramlintide,
matching placebo), 50 virtual subjects per arm.

```{r cohort}
# set.seed() seeds R's RNG. It does NOT seed rxode2's simulation RNG, and
# rxode2's streams are partitioned PER SOLVER THREAD, so this cohort is
# reproducible on this machine and different on a machine with a different
# thread count. Every assertion below is written to hold for any cohort the
# model can produce (pattern 12 of the skill's known-failure catalogue).
set.seed(20130102)
rxode2::rxSetSeed(20130102)

N_PER_ARM  <- 50L
MEAL_TIME  <- 30      # breakfast served 30 min after dose initiation (Methods)
MEAL_AMT   <- 50000   # mg of glucose, Equation 5

# Dense early sampling so NCA resolves the distribution phase, then coarser
# through the 630 min glucose window.
obs_times <- sort(unique(c(
  seq(0, 20, by = 0.5),
  seq(21, 120, by = 1),
  seq(122, 300, by = 2),
  seq(305, 630, by = 5)
)))

# OCC follows the Fang 2013 PD control-stream coding (see the model's
# covariateData$OCC): 1 placebo infusion, 2 drug infusion, 3 placebo bolus,
# 4 drug bolus. The model derives the regimen from OCC (1-2 infusion,
# 3-4 bolus) for the lag time and for the clearance IOV.
make_arm <- function(dose_ug, regimen, on_treatment, n, id_offset) {
  occ <- if (regimen == "bolus") {
    if (on_treatment) 4L else 3L
  } else {
    if (on_treatment) 2L else 1L
  }
  ids <- id_offset + seq_len(n)

  obs <- tidyr::expand_grid(id = ids, time = obs_times) |>
    dplyr::mutate(
      amt = NA_real_, evid = 0L, rate = 0,
      cmt = "central", dvid = 1L
    )

  # The meal is a zero-order input of modelled duration (rate = -2), so
  # rxode2 uses dur(intestine) = Tin * S and f(intestine) = F from the model.
  # That reproduces Equation 5, k0 = D * F / (Tin * S), exactly.
  meal <- tibble::tibble(id = ids) |>
    dplyr::mutate(
      time = MEAL_TIME, amt = MEAL_AMT, evid = 1L,
      rate = -2, cmt = "intestine", dvid = NA_integer_
    )

  rows <- dplyr::bind_rows(obs, meal)

  if (on_treatment) {
    amt_pmol <- ug_to_pmol(dose_ug)
    drug <- tibble::tibble(id = ids) |>
      dplyr::mutate(
        time = 0, amt = amt_pmol, evid = 1L,
        # Bolus regimen: modelled 4 min duration with IIV (rate = -2).
        # Infusion regimen: protocol 120 min duration, given as an explicit
        # rate so it overrides the model's dur(central).
        rate = if (regimen == "bolus") -2 else amt_pmol / 120,
        cmt = "central", dvid = NA_integer_
      )
    rows <- dplyr::bind_rows(rows, drug)
  }

  rows |>
    dplyr::mutate(
      OCC = occ,
      ON_TREATMENT = as.integer(on_treatment),
      dose_ug = dose_ug,
      regimen = regimen,
      treatment = paste(dose_ug, "ug", regimen),
      arm = paste0(dose_ug, " ug ", regimen, if (on_treatment) "" else " (placebo)")
    ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

arm_grid <- tidyr::expand_grid(
  dose_ug = c(30, 100, 300),
  regimen = c("bolus", "infusion"),
  on_treatment = c(TRUE, FALSE)
) |>
  dplyr::mutate(id_offset = (dplyr::row_number() - 1L) * N_PER_ARM)

events <- do.call(
  dplyr::bind_rows,
  lapply(seq_len(nrow(arm_grid)), function(i) {
    make_arm(
      dose_ug      = arm_grid$dose_ug[i],
      regimen      = arm_grid$regimen[i],
      on_treatment = arm_grid$on_treatment[i],
      n            = N_PER_ARM,
      id_offset    = arm_grid$id_offset[i]
    )
  })
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid", "cmt")])))
stopifnot(dplyr::n_distinct(events$id) == nrow(arm_grid) * N_PER_ARM)
```

# Simulation

`rxSolve()` is called once per arm: solving an `rxUi` is quadratic in the number
of subjects per call, so one call per arm is materially faster than one call for
the whole cohort. `useLinCmt = FALSE` is required because rxode2's automatic
ODE-to-`linCmt()` conversion corrupts the `dvid` mapping for multi-output
models.

```{r simulate}
mod <- readModelDb("Fang_2013_pramlintide")

solve_arm <- function(ev, typical = FALSE) {
  m <- if (typical) rxode2::zeroRe(rxode2::rxode(mod)) else mod
  rxode2::rxSolve(
    m, events = as.data.frame(ev),
    keep = c("OCC", "ON_TREATMENT", "dose_ug", "regimen", "treatment", "arm"),
    useLinCmt = FALSE, returnType = "data.frame"
  )
}

sim <- do.call(
  dplyr::bind_rows,
  lapply(split(events, events$arm), solve_arm)
) |>
  dplyr::as_tibble()

# One typical-value (zeroRe) subject per arm for the deterministic checks.
ev_typical <- events |> dplyr::filter(id %% N_PER_ARM == 1L)
stopifnot(dplyr::n_distinct(ev_typical$id) == nrow(arm_grid))

sim_typical <- do.call(
  dplyr::bind_rows,
  lapply(split(ev_typical, ev_typical$arm), solve_arm, typical = TRUE)
) |>
  dplyr::as_tibble()

stopifnot(nrow(sim) > 0, all(is.finite(sim$Cc)), all(is.finite(sim$Gc)))
# Solver tolerance can put the far tail marginally below zero; assert a
# magnitude, not an exact sign.
stopifnot(min(sim$Cc) > -1e-6, min(sim$Gc) > 0)
```

# Replicate published figures

## Figure 2 -- pramlintide concentration by dose group

```{r figure-2, fig.height = 6}
# Replicates Figure 2 of Fang 2013: median and 90 percent prediction interval
# of the pramlintide concentration-time profile for each dose and regimen.
sim |>
  dplyr::filter(ON_TREATMENT == 1L, time <= 300, Cc > 0) |>
  dplyr::group_by(treatment, regimen, dose_ug, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_grid(regimen ~ dose_ug, labeller = label_both) +
  scale_y_log10() +
  labs(
    x = "Time (min)", y = "Pramlintide (pmol/L)",
    title = "Figure 2 -- simulated pramlintide concentrations",
    caption = "Replicates Figure 2 of Fang 2013 (median and 5th-95th percentile)."
  )
```

## Figure 5 -- typical glucose profiles, placebo versus pramlintide

```{r figure-5, fig.height = 6}
# Replicates Figure 5 of Fang 2013: typical-value plasma glucose profiles for
# the placebo and active occasions in each dose group.
sim_typical |>
  dplyr::mutate(
    Treatment = ifelse(ON_TREATMENT == 1L, "Pramlintide", "Placebo")
  ) |>
  ggplot(aes(time, Gc, colour = Treatment, linetype = Treatment)) +
  geom_line(linewidth = 0.8) +
  facet_grid(regimen ~ dose_ug, labeller = label_both) +
  labs(
    x = "Time (min)", y = "Plasma glucose (mg/dL)",
    title = "Figure 5 -- typical postprandial glucose profiles",
    caption = paste(
      "Replicates Figure 5 of Fang 2013. Meal at 30 min;",
      "typical values with random effects zeroed."
    )
  )
```

## Figures 3 and 4 -- glucose AUCnet and Tmax by treatment

Fang 2013 computed AUCnet as the area between the glucose profile and the
individual pre-dose baseline over the 5 h following dosing, and Tmax as the time
of the peak glucose concentration over the same window.

```{r figures-3-4, fig.height = 4.5}
pd_window <- sim |>
  dplyr::filter(time <= 300) |>
  dplyr::group_by(id, arm, treatment, dose_ug, regimen, ON_TREATMENT) |>
  dplyr::summarise(
    aucnet   = {
      base <- Gc[which.min(time)]
      sum(diff(time) * (head(Gc - base, -1) + tail(Gc - base, -1)) / 2)
    },
    gluc_tmax = time[which.max(Gc)],
    .groups = "drop"
  ) |>
  dplyr::mutate(
    Treatment = ifelse(ON_TREATMENT == 1L, "Pramlintide", "Placebo")
  )

pd_summary <- pd_window |>
  dplyr::group_by(treatment, dose_ug, regimen, Treatment) |>
  dplyr::summarise(
    aucnet_mean = mean(aucnet), aucnet_sd = sd(aucnet),
    tmax_mean   = mean(gluc_tmax), tmax_sd = sd(gluc_tmax),
    .groups = "drop"
  )

ggplot(pd_summary, aes(treatment, aucnet_mean, fill = Treatment)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_errorbar(
    aes(ymin = aucnet_mean - aucnet_sd, ymax = aucnet_mean + aucnet_sd),
    position = position_dodge(0.8), width = 0.25
  ) +
  labs(
    x = NULL, y = "Glucose AUCnet (mg/dL x min)",
    title = "Figure 3 -- 5 h glucose AUCnet by treatment",
    caption = "Replicates Figure 3 of Fang 2013 (mean and SD)."
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

ggplot(pd_summary, aes(treatment, tmax_mean, fill = Treatment)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_errorbar(
    aes(ymin = tmax_mean - tmax_sd, ymax = tmax_mean + tmax_sd),
    position = position_dodge(0.8), width = 0.25
  ) +
  labs(
    x = NULL, y = "Glucose Tmax (min)",
    title = "Figure 4 -- glucose Tmax by treatment",
    caption = "Replicates Figure 4 of Fang 2013 (mean and SD)."
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

The two published pharmacodynamic findings are that pramlintide delays the
glucose Tmax and lowers the 5 h AUCnet. Both are checked below on the
**typical-value** profiles, which are deterministic and therefore reproducible
across machines.

```{r pd-checks}
pd_typ <- sim_typical |>
  dplyr::filter(time <= 300) |>
  dplyr::group_by(arm, treatment, dose_ug, regimen, ON_TREATMENT) |>
  dplyr::summarise(
    aucnet = {
      base <- Gc[which.min(time)]
      sum(diff(time) * (head(Gc - base, -1) + tail(Gc - base, -1)) / 2)
    },
    gluc_tmax = time[which.max(Gc)],
    .groups = "drop"
  )

pd_contrast <- pd_typ |>
  dplyr::select(treatment, dose_ug, regimen, ON_TREATMENT, aucnet, gluc_tmax) |>
  tidyr::pivot_wider(
    names_from = ON_TREATMENT, values_from = c(aucnet, gluc_tmax),
    names_prefix = "trt"
  ) |>
  dplyr::mutate(
    tmax_delay_min  = gluc_tmax_trt1 - gluc_tmax_trt0,
    aucnet_drop_pct = 100 * (1 - aucnet_trt1 / aucnet_trt0)
  )

knitr::kable(
  pd_contrast |>
    dplyr::select(treatment, gluc_tmax_trt0, gluc_tmax_trt1,
                  tmax_delay_min, aucnet_drop_pct) |>
    dplyr::rename(
      "Treatment" = treatment,
      "Glucose Tmax, placebo (min)" = gluc_tmax_trt0,
      "Glucose Tmax, pramlintide (min)" = gluc_tmax_trt1,
      "Tmax delay (min)" = tmax_delay_min,
      "AUCnet reduction (%)" = aucnet_drop_pct
    ),
  digits = 1,
  caption = paste(
    "Typical-value pramlintide effect on postprandial glucose.",
    "Fang 2013 reports a delayed Tmax at every dose and a decrease of at",
    "least 40 percent in mean AUCnet for all arms except the 30 ug bolus."
  )
)

is_30_bolus <- pd_contrast$treatment == "30 ug bolus"

stopifnot(
  # Everything in this chunk comes from the zeroRe (typical-value) solve, so it
  # is deterministic and an exact-shape assertion is appropriate. A
  # cohort-derived version of the same claim would need centre-and-quantile
  # bounds instead.
  #
  # Fang 2013 Results: Tmax "increased for all dosing groups compared to
  # placebo", and "the delay seems to be independent of dose".
  all(pd_contrast$tmax_delay_min > 30),
  diff(range(pd_contrast$tmax_delay_min)) < 30,
  # Fang 2013 Results: "Pramlintide dosing resulted in a decrease of at least
  # 40% in the mean 5-h integrated glucose AUCnet for all doses and dosage
  # regimens except for the 30-ug bolus group". The model reproduces that
  # split, including which single arm falls short.
  all(pd_contrast$aucnet_drop_pct[!is_30_bolus] > 40),
  pd_contrast$aucnet_drop_pct[is_30_bolus] < 40,
  # ... and it is a reduction in that arm too, just under 40 percent.
  pd_contrast$aucnet_drop_pct[is_30_bolus] > 20
)
```

The model reproduces the published split exactly: the AUCnet reduction exceeds
40 percent in five of the six arms and falls short only in the 30 ug bolus arm,
which is the one arm Fang 2013 singles out. The Tmax delay is close to
dose-independent, as the paper reports, because the gastric-emptying effect is
modelled as a maximal on/off prolongation rather than an exposure-driven one.

# PKNCA validation

Noncompartmental analysis of the simulated pramlintide concentrations, over the
0 to 300 min window the study actually sampled.

```{r pknca}
# Restrict to the active arms and to the 0-300 min window the study sampled.
# Time zero is inside this window and is retained.
sim_pk <- subset(sim, ON_TREATMENT == 1L & time <= 300)

sim_nca <- sim_pk |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time = 0 row per (id, treatment). Pramlintide is given
# intravenously with a 0.430 min lag, so the pre-dose concentration is 0.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |>
    dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

dose_df <- events |>
  dplyr::filter(evid == 1L, cmt == "central") |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(
  as.data.frame(sim_nca), Cc ~ time | treatment + id,
  concu = "pmol/L", timeu = "min"
)
dose_obj <- PKNCA::PKNCAdose(
  as.data.frame(dose_df), amt ~ time | treatment + id,
  doseu = "pmol"
)

intervals <- data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE,
  cl.obs     = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)

nca_tbl <- as.data.frame(nca_res$result)
stopifnot(nrow(nca_tbl) > 0)
```

Two internal consistency checks on the NCA machinery itself, using the
typical-value solve where the answer is known in closed form.

```{r nca-internal}
nca_median <- nca_tbl |>
  dplyr::filter(PPTESTCD %in% c("aucinf.obs", "half.life", "cl.obs")) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(value = median(PPORRES, na.rm = TRUE), .groups = "drop")

auc_med <- nca_median |> dplyr::filter(PPTESTCD == "aucinf.obs")
dose_map <- events |>
  dplyr::filter(evid == 1L, cmt == "central") |>
  dplyr::distinct(treatment, amt)

auc_chk <- auc_med |>
  dplyr::left_join(dose_map, by = "treatment") |>
  dplyr::mutate(
    auc_expected = amt / cl,                    # Dose / CL for a linear model
    pct_diff = 100 * (value - auc_expected) / auc_expected
  )

knitr::kable(
  auc_chk |>
    dplyr::select(treatment, value, auc_expected, pct_diff) |>
    dplyr::rename(
      "Treatment" = treatment,
      "Median simulated AUC0-inf (pmol*min/L)" = value,
      "Dose / CL (pmol*min/L)" = auc_expected,
      "% difference" = pct_diff
    ),
  digits = c(0, 0, 0, 2),
  caption = "Simulated AUC0-inf against the closed-form Dose/CL of the model."
)

hl_med <- nca_median |> dplyr::filter(PPTESTCD == "half.life")

stopifnot(
  # Structural: the median cohort AUC must track Dose/CL. Log-normal IIV on CL
  # means the median of AUC sits near Dose/median(CL), not the mean, so a
  # centre-of-distribution bound is the right shape here. A mis-transcribed
  # dose, volume or clearance moves this by tens of percent.
  abs(median(auc_chk$pct_diff)) < 10,
  quantile(abs(auc_chk$pct_diff), 0.9) < 20,
  # NCA half-life against the analytic terminal slope of the same model.
  abs(median(hl_med$value) - thalf_pk) / thalf_pk < 0.15
)
```

Dose proportionality. Fang 2013 fitted a power model to AUC0-inf versus dose and
reported an exponent of 0.99 with 95 percent CI (0.916, 1.25), concluding that
pramlintide PK is linear over 30 to 300 ug.

```{r dose-proportionality}
dp <- auc_chk |>
  dplyr::left_join(
    events |> dplyr::distinct(treatment, dose_ug, regimen), by = "treatment"
  )

power_b <- unname(coef(lm(log(value) ~ log(dose_ug), data = dp))[2])

cat(sprintf(
  "Simulated power coefficient b = %.4f (Fang 2013: 0.99, 95%% CI 0.916 to 1.25)\n",
  power_b
))

stopifnot(power_b > 0.916, power_b < 1.25)
```

## Comparison against published NCA

Fang 2013 Table I reports noncompartmental parameters per dose and regimen.

```{r nca-compare}
published <- tibble::tribble(
  ~treatment,         ~cmax, ~tmax, ~aucinf.obs, ~half.life, ~cl.obs,
  "30 ug bolus",       304,    4,     5410,       22.5,      1.90,
  "100 ug bolus",     1300,    6,    28200,       55.1,      0.923,
  "300 ug bolus",     3700,    7,    78900,       50.7,      0.984,
  "30 ug infusion",   67.5,   90,     7140,       19.6,      1.12,
  "100 ug infusion",   221,  110,    25400,       36.7,      1.04,
  "300 ug infusion",   827,   90,    94100,       49.8,      0.819
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by        = "treatment",
  units     = c(cmax = "pmol/L", tmax = "min", aucinf.obs = "pmol*min/L",
                half.life = "min", cl.obs = "L/min"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated vs. Fang 2013 Table I. * differs from reference by >20%.",
  align = c("l", "l", "r", "r", "r")
)
```

The exposure parameters that the model is actually built on agree closely: at
100 and 300 ug the simulated AUC0-inf is within 4 percent of Table I in three
of the four arms, and simulated CL is within 6 percent in three of the four.
The starred rows fall into four groups, each of which the paper itself
anticipates.

* **30 ug arms.** The paper states that the 30 ug profiles reached the limit of
  detection (2 pmol/L), which truncates the terminal phase and biases the
  observed AUC0-inf, CL and half-life for that group. Fang 2013 says so
  explicitly: "The shorter t1/2 at the 30 ug dose might be caused by the
  profiles reaching the limit of detection", and Table I's own CL for that arm,
  1.90 (SD 1.62) L/min, has a standard deviation of 85 percent of the mean. The
  model is a pooled linear fit and does not reproduce a truncation artefact.
* **Half-life.** The observed half-life ranges from 19.6 to 55.1 min across six
  arms of the same drug in the same subjects, which is internally inconsistent;
  the paper flags it ("we found some inconsistency in terminal half-life for the
  three dosages"). The model has one disposition and therefore one terminal
  half-life, `r sprintf("%.1f", thalf_pk)` min, which sits near the middle of
  that observed range. The `nca-internal` chunk above gates the simulated NCA
  half-life against that analytic value rather than against any single
  published arm.
* **300 ug infusion.** Observed AUC0-inf (94100) is 19 percent higher than the
  observed AUC0-inf in the same subjects after the 300 ug bolus (78900), and
  the observed Cmax (827 pmol/L) implies a clearance of about 0.77 L/min rather
  than the model's 0.955. The model, having a single clearance, predicts the
  same exposure for both regimens and matches the bolus arm closely. The
  paper's own dose-proportionality analysis pooled the regimens and found an
  exponent of 0.99, so the regimen difference in Table I is noise rather than
  structure.
* **Tmax.** Under the bolus regimen the modelled 4 min input duration plus the
  0.430 min lag places Tmax at 4.5 min for every subject, against 4, 6 and 7 min
  observed; under the infusion regimen the model peaks at the end of infusion,
  120 min, against 90 to 110 min observed. Both mismatches are the phenomenon
  the paper introduced the 4 min duration to describe in the first place: "The
  Cmax did not occur at the end of the 2-min bolus or 120-min infusion for all
  subjects", which for the long infusion the authors attribute to random
  variability.

No parameter was tuned to improve any of these comparisons.

# Assumptions and deviations

* **Molecular weight.** Fang 2013 reports doses in ug and concentrations in
  pmol/L but does not print a molecular weight. The ug-to-pmol conversion in
  this vignette uses the published pramlintide molecular weight of 3949.4 g/mol,
  which is a chemical constant and not a paper-derived value. It is
  cross-checked above against the paper's own Table I, which implies
  `r sprintf("%.0f", median(mw_check$mw_implied))` g/mol from dose, AUC and CL
  in the 100 and 300 ug arms. The model file itself contains no molecular
  weight; only the dosing unit (pmol) is declared.
* **The 60 min gate on the inhibition of endogenous glucose production.**
  Equation 6 of Fang 2013 as printed applies `I` to `kin` at all times. The
  published PD control stream instead computes
  `IF (T.LT.60) THEN N = 0 ELSE N = 1` and uses `KIN*(1.0 - N*INH)`, so the
  inhibition is inactive for the first 60 min after dose initiation. The model
  file follows the control stream, because that is the code that produced the
  Table III estimates. The gate is documented in a comment in `model()`.
* **Peripheral glucose initial condition.** The control stream sets
  `A_0(5) = 0.917 * BGLC * 85.6`. The 0.917 factor is hard-coded there, is not
  in any table, and is not implied by the distribution equilibrium of
  Equation 7 (which would give a factor of 1). It is reproduced verbatim. Its
  consequence is a small glucose decline in the first minutes after time zero,
  which the paper also observed in some placebo subjects.
* **The lag time applies only to the bolus regimen.** The published PK control
  stream assigns `ALAG1` (and the modelled duration `D1`) only inside
  `IF (REGI.EQ.0)`, that is only for the short bolus regimen. The model file
  reproduces this by multiplying the lag time by a bolus indicator derived from
  the canonical `OCC` covariate (OCC 3 and 4 are the bolus occasions). The
  modelled duration needs no indicator, because supplying an explicit `dur` or
  `rate` on the dosing record overrides `dur(central)` in rxode2, which is how
  the 120 min infusion arm is dosed here.
* **Rounded rate constants.** The control stream hard-codes the glucose
  distribution rate constants as 0.0473 and 0.0516 1/min. The model file
  computes them from the Table III values as QG/VGc =
  `r sprintf("%.5f", exp(th[["lqg"]]) / exp(th[["lvgc"]]))` and QG/VGp =
  `r sprintf("%.5f", exp(th[["lqg"]]) / exp(th[["lvgp"]]))` 1/min, which agree
  with the hard-coded values to three significant figures.
* **PK and PD random effects come from two sequential fits.** Fang 2013 fixed
  the individual empirical Bayes PK estimates when fitting the PD model, so
  Table II and Table III were not produced by one joint estimation. The single
  model file carries both sets, which is the structure the published PD control
  stream itself implements (it integrates the pramlintide PK compartments
  alongside the glucose compartments).
* **Interoccasion variability requires the `OCC` covariate.** The model uses
  indicator-multiplexed IOV etas, so `OCC` must be supplied on every record. A
  single-occasion simulation should hold `OCC` at one of 1, 2 (infusion) or
  3, 4 (bolus).
* **No covariates were retained in the published model.** Fang 2013 screened
  age, body weight and body mass index graphically and by likelihood-ratio test
  and retained none, so the model carries no demographic covariate effects.
* **Insulin is absent.** Insulin was not measured in the study, and the paper
  notes that `kout` should therefore be read as a combination of the pre-meal
  exogenous insulin effect and basal glucose removal.
* **Meal timing.** The Methods state that breakfast was served 30 min after dose
  initiation, so the simulated meal dose is placed at 30 min. The paper's
  equations are written relative to the meal dose record and do not themselves
  fix its time.
