---
title: "Marzeptacog alfa (activated) (Faraj 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Marzeptacog alfa (activated) (Faraj 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Faraj A, Knudsen T, Desai S, Neuman L, Blouse GE, Simonsson USH.
  Phase III dose selection of marzeptacog alfa (activated) informed by
  population pharmacokinetic modeling: A novel hemostatic drug. *CPT
  Pharmacometrics Syst Pharmacol.* 2022;11(12):1628-1637.
  doi:[10.1002/psp4.12872](https://doi.org/10.1002/psp4.12872)
* Description: Two-compartment population PK model for marzeptacog alfa
  (activated) (MarzAA), an activated recombinant human factor VII (rFVIIa)
  variant, after intravenous and subcutaneous dosing in adult males with
  haemophilia A or B.
* Article: <https://doi.org/10.1002/psp4.12872>
* Modality: recombinant protein (activated FVII variant), intravenous and
  subcutaneous.

Faraj 2022 pooled MarzAA concentrations from three trials and built the
model in sequential steps: intravenous data first (two-compartment
disposition), then subcutaneous absorption, then a joint re-estimation. The
final model (Results "Model development"; parameters in supplementary Table
S1; NONMEM control stream in supplementary Data S1) has

* two-compartment disposition with linear elimination (Michaelis-Menten
  elimination was dropped in the last step, dOFV = +4.3, with no change in
  the VPCs);
* subcutaneous input as a zero-order release of duration D1 into an
  absorption depot followed by first-order absorption (ka) into the central
  compartment, with a logit-transformed bioavailability F;
* allometric body-weight scaling, fixed exponents 0.75 (CL, Q) and 1
  (Vc, Vp), reference 70 kg;
* an estimated endogenous FVIIa baseline added to the drug concentration,
  because the clotting-activity assay measures endogenous FVIIa as well as
  MarzAA;
* IIV on CL and Vc (correlated), baseline, F and ka; inter-occasion
  variability (IOV) on D1, F and Vc; proportional residual error.

The model was then used to simulate candidate phase III regimens (Table 2,
Figure 2) against a target of 24 ng/mL.

# Population

46 adult males with haemophilia A or B, with or without inhibitors, from
three trials (NCT01439971, NCT04072237, NCT03407651), contributing 1225
observations (Methods "Study design and patients", Table 1). Median body
weight 75 kg (10th-90th percentile 60-94 kg; range 43-120 kg); median age
32 years (10th-90th percentile 23-52 years; range 18-62 years). Doses: single
i.v. 4.5-30 ug/kg; single s.c. 30-120 ug/kg; 60 ug/kg s.c. two or three
times 3 h apart; and daily s.c. 30 ug/kg (escalated to 60 ug/kg on
breakthrough bleeding) for 50 days. Race and region were not reported. The
same information is in the model's `population` metadata.

# Source trace

| Element | Value | Source |
|---|---|---|
| Structure: 2-cmt, zero-order into depot then first-order ka | -- | Results "Model development" final paragraph; Data S1 `$DES` |
| Baseline added to drug concentration | IPRED = BASE + A(2)/Vc | Methods "Model building"; Data S1 `$ERROR` |
| Allometry, reference 70 kg | CL, Q: 0.75; Vc, Vp: 1 | Methods "Model building"; Data S1 `$PK` |
| `lcl` | 1085 mL/h | Table S1 |
| `lvc` | 3393 mL | Table S1 |
| `lvp` | 619 mL | Table S1 |
| `lq` | 213 mL/h | Table S1 |
| `lrbase` | 0.96 ng/mL | Table S1 |
| `lka` | 0.05 1/h | Table S1 |
| `logitfdepot` | logit(0.26) | Table S1; Data S1 `PHI = LOG(TVF1/(1-TVF1))` |
| `ld1` | 0.875 h | Table S1 |
| `etalcl`, `etalvc` | 28%, 26%, correlation 0.92 | Table S1 |
| `etalrbase` | 56% | Table S1 |
| `etalogitfdepot` | 54% (logit scale) | Table S1 |
| `etalka` | 30% | Table S1 |
| `etaiov_ld1_*` | 61% | Table S1; Data S1 IOV block |
| `etaiov_logitfdepot_*` | 21% (logit scale) | Table S1; Data S1 IOV block |
| `etaiov_lvc_*` | 45% | Table S1; Data S1 IOV block |
| `propSd` | 29% | Table S1; Data S1 `Y = IPRED*EXP(EPS(1))` |

# Virtual cohort and simulation

Faraj 2022 simulated 1000 subjects per regimen with body weight drawn from a
truncated normal distribution (bounds 40-105 kg) using the mean and SD of the
analysis data set. The mean and SD are not printed; they are approximated
here from the median (75 kg) and 10th-90th percentiles (60-94 kg) as mean
76 kg, SD 13.3 kg. Each arm uses 200 subjects. Subcutaneous doses go to the
`depot` compartment with `rate = -2` so that the modelled zero-order duration
D1 applies.

All simulated subjects take `OCC = 2`, i.e. one subcutaneous occasion with
its IOV draw. The paper does not say whether its simulations included IOV;
with IOV the simulated Tmax 80% interval matches Table 2 (5.3-11.1 h for a
single dose), while without IOV it is about half as wide, so IOV is kept.

```{r cohort}
rxode2::rxSetSeed(20221216)
set.seed(20221216)
n_per_arm <- 200

draw_wt <- function(n) {
  x <- rnorm(5 * n, mean = 76, sd = 13.3)
  x[x >= 40 & x <= 105][seq_len(n)]
}

regimens <- tibble::tribble(
  ~treatment,             ~dose, ~times,
  "60 ug/kg single",      60,    list(0),
  "90 ug/kg single",      90,    list(0),
  "120 ug/kg single",     120,   list(0),
  "60 ug/kg x2 (0, 3 h)", 60,    list(c(0, 3)),
  "60 ug/kg x3 (0, 3, 6 h)", 60, list(c(0, 3, 6))
)

obs_times <- sort(unique(c(seq(0, 24, by = 0.25))))

make_arm <- function(k) {
  r <- regimens[k, ]
  wt <- draw_wt(n_per_arm)
  ids <- (k - 1) * n_per_arm + seq_len(n_per_arm)
  dose_times <- r$times[[1]][[1]]
  doses <- expand.grid(i = seq_len(n_per_arm), time = dose_times)
  doses <- data.frame(
    id = ids[doses$i], time = doses$time, evid = 1L,
    amt = r$dose * wt[doses$i], cmt = "depot", rate = -2,
    WT = wt[doses$i]
  )
  obs <- expand.grid(i = seq_len(n_per_arm), time = obs_times)
  obs <- data.frame(
    id = ids[obs$i], time = obs$time, evid = 0L,
    amt = 0, cmt = "central", rate = 0,
    WT = wt[obs$i]
  )
  out <- rbind(doses, obs)
  out$OCC <- 2L
  out$treatment <- r$treatment
  out
}

events <- do.call(rbind, lapply(seq_len(nrow(regimens)), make_arm))
events <- events[order(events$id, events$time, -events$evid), ]
```

```{r simulate}
mod <- readModelDb("Faraj_2022_marzeptacogAlfa")
sim <- rxode2::rxSolve(
  mod, events,
  keep = c("treatment", "WT"),
  returnType = "data.frame"
)
sim <- sim |>
  mutate(treatment = factor(treatment, levels = regimens$treatment))
stopifnot(!anyNA(sim$Cc))
```

`Cc` is the individual prediction without residual error and, like the
paper's predictions, includes the endogenous FVIIa baseline.

# Concentration-time profiles

```{r profiles, fig.width = 7, fig.height = 5}
sim |>
  group_by(treatment, time) |>
  summarise(
    med = median(Cc), lo = quantile(Cc, 0.1), hi = quantile(Cc, 0.9),
    .groups = "drop"
  ) |>
  ggplot(aes(time, med)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.25) +
  geom_line() +
  geom_hline(yintercept = c(24, 120), linetype = "dashed") +
  facet_wrap(~treatment) +
  scale_y_log10() +
  labs(
    x = "Time after first dose (h)", y = "FVIIa activity (ng/mL)",
    caption = "Median and 80% prediction interval; dashed lines: 24-120 ng/mL target range."
  )
```

# Percentage of the population above target (Figure 2)

```{r above-target, fig.width = 7, fig.height = 4}
above <- sim |>
  group_by(treatment, time) |>
  summarise(pct = 100 * mean(Cc >= 24), .groups = "drop")

ggplot(above, aes(time, pct, colour = treatment)) +
  geom_line() +
  geom_hline(yintercept = 90, linetype = "dashed") +
  labs(
    x = "Time after first dose (h)", y = "Population above 24 ng/mL (%)",
    colour = NULL,
    caption = "Replicates Figure 2 of Faraj 2022."
  )
```

The paper reports the percentages at selected times in Results "Clinical
trial simulations" and the Discussion.

```{r above-target-table}
ref_above <- tibble::tribble(
  ~treatment,              ~time, ~paper,
  "60 ug/kg single",       3,     51,
  "90 ug/kg single",       3,     74,
  "120 ug/kg single",      3,     86,
  "60 ug/kg single",       6,     70,
  "90 ug/kg single",       6,     87,
  "120 ug/kg single",      6,     94,
  "60 ug/kg single",       24,    33,
  "90 ug/kg single",       24,    63,
  "120 ug/kg single",      24,    78,
  "60 ug/kg x2 (0, 3 h)",  6,     90,
  "60 ug/kg x2 (0, 3 h)",  24,    81,
  "60 ug/kg x3 (0, 3, 6 h)", 9,   96
)
cmp_above <- ref_above |>
  left_join(above |> mutate(treatment = as.character(treatment)),
            by = c("treatment", "time")) |>
  mutate(diff = pct - paper)
cmp_above |>
  dplyr::rename(
    Regimen = treatment, "Time (h)" = time, "Paper (%)" = paper,
    "Simulated (%)" = pct, "Difference (points)" = diff
  ) |>
  knitr::kable(digits = 0)
```

The simulated percentages agree with the paper within a few points. With
200 subjects per arm the binomial standard error is up to 3.5 points, so
differences of 5-7 points at a single time point are within sampling noise.

```{r above-target-check}
# 200 subjects per arm gives a binomial SE of up to 3.5 points per time point;
# a transcription error in F, ka, CL or a volume moves these by tens of points.
stopifnot(
  abs(median(cmp_above$diff)) < 5,
  quantile(abs(cmp_above$diff), 0.9) < 12
)
```

# NCA (PKNCA) and comparison with Table 2

Table 2 reports the median and 80% prediction interval of Cmax, Tmax, Cmin
and AUC over 0-24 h. After the subcutaneous doses the concentration rises
until Tmax and then declines, so the paper's Cmin over 24 h is the
concentration at 24 h; it is compared with PKNCA's `clast.obs` (last
observation, at 24 h). The 0-24 h minimum including the pre-dose time point
would be the endogenous baseline instead.

```{r pknca}
conc <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)
dose <- events |>
  dplyr::filter(evid == 1) |>
  select(id, time, amt, treatment) |>
  mutate(treatment = factor(treatment, levels = regimens$treatment))

conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose, amt ~ time | treatment + id)
intervals <- data.frame(
  start = 0, end = 24,
  cmax = TRUE, tmax = TRUE, clast.obs = TRUE, auclast = TRUE
)
nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r nca-comparison}
reference <- tibble::tribble(
  ~treatment,                ~cmax, ~tmax, ~clast.obs, ~auclast,
  "60 ug/kg single",         34.5,  7.8,   19.3,       604,
  "90 ug/kg single",         51.1,  7.8,   28.3,       905.3,
  "120 ug/kg single",        67.7,  7.8,   37.4,       1207.1,
  "60 ug/kg x2 (0, 3 h)",    68.7,  9.5,   40.4,       1166,
  "60 ug/kg x3 (0, 3, 6 h)", 106.1, 11.6,  66.7,       1715
)
tbl <- nlmixr2lib::ncaComparisonTable(
  simulated = nca,
  reference = reference,
  by = "treatment",
  units = c(cmax = "ng/mL", tmax = "h", clast.obs = "ng/mL", auclast = "h*ng/mL")
)
knitr::kable(tbl)
if (!is.null(attr(tbl, "footnote"))) cat(attr(tbl, "footnote"))
```

```{r nca-check}
pct <- as.numeric(gsub("[^0-9.-]", "", tbl[["% diff"]]))
stopifnot(!anyNA(pct))
# Medians of 200 subjects; a mis-transcribed F, ka, CL or volume moves the
# whole distribution by tens of percent.
stopifnot(
  abs(median(pct)) < 7,
  quantile(abs(pct), 0.9) < 15
)
```

All simulated medians are within 10% of Table 2. The two repeated-dose
Cmax rows run about 8% below the paper; Tmax, the 24 h concentration and
AUC0-24 for the same regimens agree within 4%.

The 80% prediction intervals are compared for the single 60 ug/kg dose,
where Table 2 prints Cmax 17.6-68.5 ng/mL, Tmax 5.3-11.1 h and AUC0-24
310-1190 h*ng/mL.

```{r pi-60}
nca$result |>
  dplyr::filter(treatment == "60 ug/kg single",
                PPTESTCD %in% c("cmax", "tmax", "clast.obs", "auclast")) |>
  group_by(PPTESTCD) |>
  summarise(
    median = median(PPORRES), p10 = quantile(PPORRES, 0.1),
    p90 = quantile(PPORRES, 0.9), .groups = "drop"
  ) |>
  dplyr::rename(Parameter = PPTESTCD, Median = median, "10th pct" = p10, "90th pct" = p90) |>
  knitr::kable(digits = 1)
```

# Intravenous dosing

A typical-value profile after the 18 ug/kg i.v. dose used in two of the
trials, beside the 60 ug/kg subcutaneous dose, for a 70 kg subject. The
subcutaneous profile is absorption-limited (ka = 0.05 1/h is well below the
elimination rate constant CL/Vc = 0.32 1/h).

```{r typical, fig.width = 6, fig.height = 4}
typ_ev <- rbind(
  data.frame(id = 1L, time = 0, evid = 1L, amt = 18 * 70, cmt = "central", rate = 0),
  data.frame(id = 1L, time = seq(0, 48, by = 0.25), evid = 0L, amt = 0, cmt = "central", rate = 0),
  data.frame(id = 2L, time = 0, evid = 1L, amt = 60 * 70, cmt = "depot", rate = -2),
  data.frame(id = 2L, time = seq(0, 48, by = 0.25), evid = 0L, amt = 0, cmt = "central", rate = 0)
)
typ_ev$WT <- 70
typ_ev$OCC <- 1L
typ <- rxode2::rxSolve(rxode2::zeroRe(mod), typ_ev, returnType = "data.frame") |>
  mutate(route = ifelse(id == 1, "18 ug/kg i.v.", "60 ug/kg s.c."))
stopifnot(!anyNA(typ$Cc))
ggplot(typ, aes(time, Cc, colour = route)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Time (h)", y = "FVIIa activity (ng/mL)", colour = NULL)
```

# Assumptions and deviations

* **Random-effect scale.** Table S1 prints all random effects as
  percentages. They are read as 100 x the SD of the eta, so each variance is
  (value/100)^2. This is the only reading available for the logit-scale F
  etas, whose percentages cannot be log-normal CVs; for the log-normal
  parameters the difference from the log-normal CV conversion is small
  (for example 0.314 vs 0.272 for the 56% baseline IIV).
* **CL-Vc correlation.** Table S1 "Covariance between CL and V (%)" = 92 is
  read as a correlation of 0.92 (covariance 0.92 x 0.28 x 0.26 = 0.0670).
* **Initial versus final estimates.** The `$THETA` and `$OMEGA` values in
  Data S1 are initial estimates (for example CL 1300 mL/h, Vp 988 mL) and
  differ from the final values in Table S1, which are used here. Data S1 is
  used only for the model structure.
* **Fixed-to-zero terms omitted.** Data S1 carries an absorption lag time
  fixed at 0 and IIV on Vp and D1 fixed at 0; these are left out. Q and Vp
  have no IIV.
* **Occasions.** Data S1 indexes the IOV on the data column REGI: REGI = 1
  has no IOV, and REGI = 2-9 each have their own IOV eta with a shared
  variance. The model's `OCC` column takes the REGI value. The paper does
  not define REGI beyond describing IOV for subcutaneous data obtained on
  more than one occasion per subject. Use `OCC = 1` for no IOV, or 2-9 for
  separate occasions.
* **Simulation cohort.** The paper simulated 1000 subjects per regimen with
  weights from a truncated normal distribution using the data-set mean and
  SD, which are not printed; this vignette uses 200 per regimen with mean
  76 kg and SD 13.3 kg, approximated from the printed median and 10th-90th
  percentiles. Whether the paper's simulations included IOV is not stated;
  including it reproduces the Table 2 Tmax interval.
* **Age.** Age on ka was statistically significant but not clinically
  relevant and was not retained in the final model; no coefficient is
  printed, so it is recorded only in `covariatesDataExcluded`.
* **Endogenous baseline.** `Cc` includes the endogenous FVIIa baseline
  (typical 0.96 ng/mL). To get the MarzAA contribution alone, subtract
  `rbase` or set `lrbase` to a very small value.
