---
title: "Gamma-hydroxybutyric acid, rat (Felmlee 2010)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gamma-hydroxybutyric acid, rat (Felmlee 2010)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5, dpi = 96
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
rxode2::rxSetSeed(20260903)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Felmlee_2010_ghb_rat"))$reference`
* Description: `r rxode2::rxode(readModelDb("Felmlee_2010_ghb_rat"))$description`
* Article: <https://doi.org/10.1208/s12248-010-9197-x>

Felmlee and colleagues built the first mechanistic toxicokinetic (TK) model for
gamma-hydroxybutyric acid (GHB) that carries *both* of the drug's nonlinearities
at once: capacity-limited metabolism and saturable active renal reabsorption.
The motivating question is therapeutic. GHB overdose has no antidote; because
reabsorption from the proximal tubule is monocarboxylate-transporter (MCT)
mediated and saturable, inhibiting it should increase renal clearance and lower
plasma exposure. The paper uses the fitted model to simulate that intervention
and then corroborates the prediction experimentally with L-lactate, a known MCT
inhibitor.

```{r load-model}
mod <- rxode2::rxode(readModelDb("Felmlee_2010_ghb_rat"))
```

# Population

```{r population, echo = FALSE}
pop <- mod$meta$population
tibble::tibble(
  Field = c("Species", "N", "Weight", "Doses", "Sampling", "Region"),
  Value = c(
    pop$species,
    paste0(pop$n_subjects, " (7-10 per dose group)"),
    pop$weight_range,
    "200, 400, 600 or 1,000 mg/kg, single IV bolus (jugular vein)",
    "Plasma 0-360 min (12 points); urine 0-60, 60-120, 120-240, 240-360 min",
    pop$regions
  )
) |>
  knitr::kable()
```

Male Sprague-Dawley rats (Harlan), 280-320 g, cannulated in the right jugular
vein and housed in metabolic cages. GHB in plasma and urine was measured by
LC/MS/MS. Plasma concentration and cumulative urinary excretion from all four
doses were fitted **simultaneously** in NONMEM VI (level 1.1) using FOCE and the
ADVAN9 differential-equation solver (source paper Methods, "Population TK
Modeling").

A separate interaction study (GHB 600 mg/kg alone, N = 3, versus GHB 600 mg/kg
plus L-lactate 330 mg/kg IV bolus with a 121 mg/kg/h IV infusion, N = 5) was run
to confirm the inhibition simulations; it did not contribute to the model fit.

# Structural model

The model is *hybrid physiological* (semi-PBPK): several volumes and flows are
fixed to rat physiological values and the remaining nonlinear parameters are
estimated.

* `central` (plasma) exchanges with `peripheral1` (fast tissue) and
  `peripheral2` (slow tissue) at distributional clearances CLD and CLD2.
* Capacity-limited **metabolic** elimination (Michaelis-Menten) acts on
  `peripheral1`, not on plasma.
* `central` exchanges with `kidney` at renal blood flow QR, and drug is filtered
  out of `kidney` at GFR into `ulf1`, the proximal-tubule ultrafiltrate.
* Saturable **active reabsorption** (MCT-mediated, Michaelis-Menten) returns
  drug from `ulf1` to `kidney`. Reabsorption occurs from the proximal tubule
  only.
* Urine flow UF carries the surviving filtrate `ulf1` -> `ulf2` (distal tubule)
  -> `urine`. The distal compartment exists to reproduce the delay between
  filtration at the glomerulus and the appearance of drug in voided urine.

The two ultrafiltrate states are declared through the documented
`paper_specific_compartments` escape hatch, because `ulf1` / `ulf2` are specific
to renal-tubule models rather than library-wide canonical compartments.

## Transport-inhibition extension

Source paper Eqs. 11-13 modify the reabsorption rate law for a steady-state
inhibitor at ratio `R = [I] / Ki`. The packaged model encodes all three
mechanisms in one branch-free expression using two multiplier covariates:

```
reabsorption <- vmax_reab /
  (km_reab * (1 + INH_MCT_KM_RATIO) + Culf1 * (1 + INH_MCT_CONC_RATIO))
```

```{r mechanism-table, echo = FALSE}
tibble::tibble(
  Mechanism = c("None (fitted model, Eq. 4)", "Competitive (Eq. 11)",
                "Noncompetitive (Eq. 12)", "Uncompetitive (Eq. 13)"),
  INH_MCT_KM_RATIO = c("0", "R", "R", "0"),
  INH_MCT_CONC_RATIO = c("0", "0", "R", "R")
) |>
  knitr::kable()
```

Both covariates default to 0, so a user who never sets them recovers the fitted
model exactly. Only the competitive case is tabulated by the paper (Table II and
Fig. 3); the noncompetitive and uncompetitive results are reported as "data not
shown".

# Source trace

Every equation and every `ini()` value, with its location in the source paper.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Item, ~Source, ~Value,
  "d/dt(central)", "Eq. 1", "-(QR + CLD + CLD2)*Cc + QR*Ck + CLD*Cp1 + CLD2*Cp2",
  "d/dt(peripheral1)", "Eq. 2", "CLD*Cc - (CLD + Vmax,m/(Km,m + Cp1))*Cp1",
  "d/dt(peripheral2)", "Eq. 3", "CLD2*Cc - CLD2*Cp2",
  "Reabsorption", "Eq. 4", "Vmax,R / (Km,R + Culf1)   [a clearance]",
  "d/dt(kidney)", "Eq. 5", "QR*Cc - (QR + GFR)*Ck + Reabsorption*Culf1",
  "d/dt(ulf1)", "Eq. 6", "GFR*Ck - UF*Culf1 - Reabsorption*Culf1",
  "d/dt(ulf2)", "Eq. 7", "UF*Culf1 - UF*Culf2",
  "d/dt(urine)", "Eq. 8", "UF*Culf2",
  "Plasma residual error", "Eq. 9", "Y = Log(Aplasma/Vplasma) + eps_plasma",
  "Urine residual error", "Eq. 10", "Y = Log(Aurine) + eps_urine",
  "Competitive inhibition", "Eq. 11", "Km,R * (1 + R)",
  "Noncompetitive inhibition", "Eq. 12", "Km,R * (1 + R), Culf1 * (1 + R)",
  "Uncompetitive inhibition", "Eq. 13", "Culf1 * (1 + R)",
  "Time-averaged renal clearance", "Eq. 14", "CLR = Ae,inf / AUC_plasma",
  "lvc (Vplasma)", "Table I", "10.5 mL, FIXED (footnote a)",
  "lvp (Vtissue1)", "Table I", "75.9 mL (16% CV)",
  "lvp2 (Vtissue2)", "Table I", "26.8 mL",
  "lv_kidney (Vkidney)", "Table I", "4.0 mL, FIXED (footnote a)",
  "lv_ulf1 (Vulf1)", "Table I", "3.0 mL, FIXED (footnote a)",
  "lv_ulf2 (Vulf2)", "Table I", "1.0 mL, FIXED (footnote a)",
  "lq (CLD)", "Table I", "26.9 mL/min",
  "lq2 (CLD2)", "Table I", "3.07 mL/min",
  "lq_kidney (QR)", "Table I", "12.5 mL/min, FIXED (footnote a)",
  "lvmax (Vmax,m)", "Table I", "0.581 mg/min",
  "lkm (Km,m)", "Table I", "0.054 mg/mL",
  "lgfr (GFR)", "Table I", "10 mL/min/kg (see Errata)",
  "lvmax_reab (Vmax,R)", "Table I", "2.34 mg/min",
  "lkm_reab (Km,R)", "Table I", "0.46 mg/mL",
  "luf (UF)", "Table I", "0.1 mL/min, FIXED (footnote a) (114% CV)",
  "expSd (eps_plasma)", "Table I", "2.5%",
  "expSd_urine (eps_urine)", "Table I", "46%"
) |>
  knitr::kable()
```

Between-subject variability is exponential (`P_i = theta * exp(eta_i)`, source
paper "Structural, Parameter Variability, and Observational Models"), reported
in Table I as CV% on exactly two parameters. The packaged model uses
`omega^2 = log(1 + CV^2)`, the same convention as the companion rat GHB model
already in the library, `Fung_2008_butanediol_rat.R`, whose variability
reporting this paper explicitly says it followed (source paper reference 26).

# Simulation setup

```{r sim-setup}
WT_KG   <- 0.3                       # 280-320 g study rats; see Errata
DOSES   <- c(200, 400, 600, 1000)    # mg/kg
R_LEVEL <- c(0, 1, 10, 100)          # [I]/Ki for competitive inhibition
NSUB    <- 200                       # per arm; the paper simulated 1,000

# Route A event table (ODE-state cmt + explicit dvid). This model declares two
# endpoints (Cc and urine), so every observation row must say which endpoint it
# belongs to: dvid 1 = plasma concentration, dvid 2 = cumulative urine amount.
build_events <- function(dose_mgkg, r_km = 0, r_conc = 0,
                         tgrid = seq(0, 360, by = 1), both_endpoints = FALSE) {
  obs <- data.frame(
    time = tgrid, amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L
  )
  if (both_endpoints) {
    obs <- rbind(obs, transform(obs, dvid = 2L))
  }
  ev <- rbind(
    data.frame(
      time = 0, amt = dose_mgkg * WT_KG, evid = 1L,
      cmt = "central", dvid = NA_integer_
    ),
    obs
  )
  ev <- ev[order(ev$time, -ev$evid), ]
  ev$WT <- WT_KG
  ev$INH_MCT_KM_RATIO <- r_km
  ev$INH_MCT_CONC_RATIO <- r_conc
  ev
}

# Trapezoidal AUC, matching the paper's WinNonLin linear-trapezoidal AUC.
trap_auc <- function(time, conc) {
  sum(diff(time) * (utils::head(conc, -1) + utils::tail(conc, -1)) / 2)
}
```

# Reproducing Table II: plasma AUC under competitive inhibition

Table II is the paper's central quantitative result: plasma AUC and
time-averaged renal clearance for each of the four doses, with no inhibitor and
at `R` = 1, 10 and 100. The AUC column is the load-bearing check on the
transcription, because it depends on every ODE, every Table I parameter, the
dose/body-weight bridge and the competitive-inhibition rate law at once.

Sixteen typical-value (no random effects) solves, one per dose x `R` cell:

```{r table2-auc}
scen <- tidyr::expand_grid(dose = DOSES, R = R_LEVEL) |>
  dplyr::mutate(id = dplyr::row_number())

ev_all <- do.call(rbind, lapply(seq_len(nrow(scen)), function(i) {
  e <- build_events(scen$dose[i], r_km = scen$R[i], tgrid = seq(0, 360, by = 0.5))
  e$id <- scen$id[i]
  e
}))

sim_tv <- rxode2::rxSolve(
  mod, ev_all, returnType = "data.frame",
  omega = NA, sigma = NA, addDosing = FALSE
)

auc_tv <- sim_tv |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(id) |>
  dplyr::summarise(
    auc = trap_auc(time, Cc),
    ae  = max(urine),
    .groups = "drop"
  ) |>
  dplyr::left_join(scen, by = "id")
```

```{r table2-compare, echo = FALSE}
published_auc <- tidyr::expand_grid(dose = DOSES, R = R_LEVEL) |>
  dplyr::mutate(published = c(
    29.3, 27.4, 21.0, 14.4,
    83.0, 76.2, 56.6, 37.5,
    133.3, 124.6, 96.1, 63.6,
    208.2, 198.8, 165.2, 117.1
  ))

chk_auc <- auc_tv |>
  dplyr::left_join(published_auc, by = c("dose", "R")) |>
  dplyr::mutate(pct_diff = 100 * (auc - published) / published)

chk_auc |>
  dplyr::transmute(
    `GHB dose (mg/kg)` = dose,
    `R = [I]/Ki` = R,
    `Published AUC` = published,
    `Simulated AUC` = round(auc, 1),
    `% diff` = round(pct_diff, 1)
  ) |>
  knitr::kable(
    caption = "Replicates Table II of Felmlee 2010 (plasma AUC, min*mg/mL, competitive inhibition)."
  )
```

```{r table2-assert}
# Deterministic typical-value solves: no random draws on either side, so a
# tight bound is reproducible across rxode2 builds.
stopifnot(
  median(abs(chk_auc$pct_diff)) < 6,
  max(abs(chk_auc$pct_diff)) < 12
)
```

All sixteen cells agree with the published values, and the agreement is tightest
at the two highest doses, which carry the most informative data. The AUC falls
monotonically with increasing `R` at every dose, reproducing the paper's
headline claim that inhibiting reabsorption lowers GHB exposure.

# Reproducing Figure 3: profiles under competitive inhibition

```{r fig3, fig.height = 6, fig.cap = "Replicates Figure 3 of Felmlee 2010: simulated plasma GHB concentration-time profiles in the absence and presence of a competitive inhibitor of renal reabsorption."}
sim_tv |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::left_join(scen, by = "id") |>
  dplyr::mutate(
    dose_lab = factor(paste0(format(dose, big.mark = ","), " mg/kg"),
                      levels = paste0(format(DOSES, big.mark = ","), " mg/kg")),
    R_lab = factor(R, levels = R_LEVEL,
                   labels = c("no inhibitor", "R = 1", "R = 10", "R = 100"))
  ) |>
  ggplot2::ggplot(ggplot2::aes(time, Cc, colour = R_lab, linetype = R_lab)) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::facet_wrap(~dose_lab, scales = "free_y") +
  ggplot2::scale_y_log10() +
  ggplot2::labs(
    x = "Time (min)", y = "GHB plasma concentration (mg/mL)",
    colour = NULL, linetype = NULL
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "top")
```

# Reproducing Figure 2: population variability

Figure 2 shows, for each dose, the mean and the 10th and 90th percentiles of
1,000 simulated profiles against the observed data. Two details of that figure
govern how it must be reproduced:

1. The bands are computed from *simulated observations*, so they include the
   residual error of Eqs. 9-10 (2.5% for plasma, 46% for urine), not only
   between-subject variability. In `rxSolve()` output the residual-error column
   is `sim`; the `Cc` / `ipredSim` columns are individual predictions with no
   residual error.
2. Urinary excretion carries a 114% CV on urine flow, which makes the
   distribution of excreted amount strongly right-skewed. The *mean* line
   therefore sits well above the median, which is what the published figure
   shows.

```{r fig2-sim}
cohort <- do.call(rbind, lapply(DOSES, function(d) {
  ev <- build_events(d, tgrid = seq(0, 360, by = 5), both_endpoints = TRUE)
  rxode2::rxSolve(
    mod, ev, nSub = NSUB, returnType = "data.frame",
    addDosing = FALSE, keep = "dvid"
  ) |>
    dplyr::mutate(dose = d)
}))

bands <- cohort |>
  dplyr::mutate(
    endpoint = ifelse(dvid == 1L, "Plasma (mg/mL)", "Cumulative urine (mg)")
  ) |>
  dplyr::group_by(dose, endpoint, time) |>
  dplyr::summarise(
    mean = mean(sim),
    p10  = quantile(sim, 0.10),
    p90  = quantile(sim, 0.90),
    .groups = "drop"
  )
```

```{r fig2-plot, fig.height = 6.5, fig.cap = "Replicates Figure 2 of Felmlee 2010: mean (solid) with 10th and 90th percentiles (dashed) of simulated plasma concentration and cumulative urinary excretion."}
bands |>
  dplyr::mutate(
    dose_lab = factor(paste0(format(dose, big.mark = ","), " mg/kg"),
                      levels = paste0(format(DOSES, big.mark = ","), " mg/kg"))
  ) |>
  ggplot2::ggplot(ggplot2::aes(time)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = p10, ymax = p90), alpha = 0.15) +
  ggplot2::geom_line(ggplot2::aes(y = mean), linewidth = 0.7) +
  ggplot2::geom_line(ggplot2::aes(y = p10), linetype = "dashed", linewidth = 0.3) +
  ggplot2::geom_line(ggplot2::aes(y = p90), linetype = "dashed", linewidth = 0.3) +
  ggplot2::facet_grid(endpoint ~ dose_lab, scales = "free_y") +
  ggplot2::labs(x = "Time (min)", y = NULL) +
  ggplot2::theme_bw()
```

The published Figure 2 mean cumulative urinary excretion at 360 min reads
approximately 5.7, 27.5, 78 and 170 mg at 200, 400, 600 and 1,000 mg/kg. Those
values are also recoverable arithmetically from Table II, since Eq. 14 gives
`Ae,inf = CLR * AUC`: 0.197 x 29.3 = 5.8, 0.331 x 83.0 = 27.5,
0.540 x 133.3 = 72.0 and 0.823 x 208.2 = 171.3 mg. Three of the four agree
closely with the figure; the 600 mg/kg point differs by about 8% (roughly 78 mg
read off the figure against 72.0 mg from the table), which is within the
precision of reading a value off a log-scaled panel. That correspondence is what
identifies both the figure and the table as reporting *population means
including residual error* rather than typical values.

```{r fig2-check}
ae360 <- cohort |>
  dplyr::filter(dvid == 2L, time == 360) |>
  dplyr::group_by(dose) |>
  dplyr::summarise(mean_ae = mean(sim), .groups = "drop") |>
  dplyr::mutate(
    published_ae = c(29.3, 83.0, 133.3, 208.2) * c(0.197, 0.331, 0.540, 0.823),
    pct_diff = 100 * (mean_ae - published_ae) / published_ae
  )

knitr::kable(
  dplyr::transmute(
    ae360,
    `GHB dose (mg/kg)` = dose,
    `Published Ae,inf (mg)` = round(published_ae, 1),
    `Simulated mean Ae (mg)` = round(mean_ae, 1),
    `% diff` = round(pct_diff, 1)
  ),
  caption = "Cumulative urinary excretion at 360 min, back-calculated from Table II via Eq. 14."
)

# A cohort mean over a 114%-CV parameter is a heavy-tailed statistic, so this
# assertion bounds the centre of the comparison rather than any single dose.
stopifnot(median(abs(ae360$pct_diff)) < 40)
```

The simulated cohort means fall short of the published ones at the lower doses,
most at 200 mg/kg, and close the gap as dose rises. This is a reconstruction gap
in a derived metric, not a transcription error, and it is discussed under Errata
below. Note that the cohort mean of a 114%-CV parameter is a heavy-tailed
statistic, so the per-dose percentages move with the random draw; only the
centre of the comparison is asserted above.

# PKNCA validation

Noncompartmental analysis of the typical-value plasma profiles, on the same
0-360 min window the paper sampled. The NCA is run on the typical-value solve
rather than the full-variability cohort: with a 114% CV on urine flow, terminal
slopes across a simulated cohort are heterogeneous enough that a pooled
half-life is dominated by which subjects happened to draw a slow tubular flow.

```{r pknca}
# Select the no-inhibitor arms by id up front, so the only row-dropping filter
# in the PKNCA chain below is the mandatory !is.na() one.
tv_ids <- scen$id[scen$R == 0]

nca_conc <- sim_tv[sim_tv$id %in% tv_ids, ] |>
  dplyr::left_join(scen, by = "id") |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::transmute(
    id = as.integer(id),
    treatment = paste0(format(dose, big.mark = ","), " mg/kg"),
    time, Cc
  )

# PKNCA warns "AUC range starting (0) before the first measurement" once per
# subject if any profile lacks a time-zero record, so assert it rather than
# discovering it in the render log.
stopifnot(all(table(nca_conc$treatment[nca_conc$time == 0]) == 1))

nca_dose <- nca_conc |>
  dplyr::distinct(id, treatment) |>
  dplyr::mutate(time = 0, amt = DOSES * WT_KG)

# Additive grouping (| treatment + id) on BOTH objects: PKNCAdose rejects a
# nested slash formula. `dose` and `route` are reserved PKNCA column names, so
# the grouping column is `treatment`.
o_conc <- PKNCA::PKNCAconc(nca_conc, Cc ~ time | treatment + id,
                           concu = "mg/mL", timeu = "min")
o_dose <- PKNCA::PKNCAdose(nca_dose, amt ~ time | treatment + id, doseu = "mg")

intervals <- data.frame(
  start = 0, end = 360,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, clast.obs = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals))
nca_sum <- as.data.frame(nca_res)
```

```{r pknca-table, echo = FALSE}
nca_sum |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::mutate(dplyr::across(where(is.numeric), ~ signif(.x, 3))) |>
  dplyr::rename(
    "Treatment" = treatment,
    "Cmax (mg/mL)" = cmax,
    "Tmax (min)" = tmax,
    "AUClast (min*mg/mL)" = auclast,
    "Clast (mg/mL)" = clast.obs
  ) |>
  knitr::kable(caption = "PKNCA noncompartmental parameters from the typical-value plasma profiles.")
```

## Comparison against the published AUC

The paper reports no Cmax / Tmax / half-life table, so the AUC column of Table II
is the only published NCA quantity available for a side-by-side comparison. It
was itself computed by NCA in WinNonLin 5.2 (source paper Methods, "Statistical
and Non-Compartmental Analyses"; the paper does not state which trapezoidal rule
was used), so `auclast` over 0-360 min is the matching parameter.

```{r nca-comparison}
ref_nca <- data.frame(
  treatment = paste0(format(DOSES, big.mark = ","), " mg/kg"),
  auclast = c(29.3, 83.0, 133.3, 208.2)
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_sum,
  reference = ref_nca,
  by = "treatment",
  params = "auclast",
  units = c(auclast = "min*mg/mL")
)
knitr::kable(cmp, caption = "Simulated versus published plasma AUC (Felmlee 2010 Table II, GHB alone).")
attr(cmp, "footnote")
```

No row exceeds the 20% flagging tolerance.

# The L-lactate interaction study (Table III)

Table III reports the experimental confirmation: GHB 600 mg/kg alone versus the
same dose co-administered with L-lactate. The paper notes that the observed
changes resemble the simulated `R = 10` case.

```{r table3, echo = FALSE}
auc600 <- chk_auc |> dplyr::filter(dose == 600)

tibble::tibble(
  Arm = c("GHB alone (N = 3), observed", "GHB + L-lactate (N = 5), observed",
          "GHB alone, simulated", "GHB + competitive inhibitor R = 10, simulated"),
  `AUC (min*mg/mL)` = c(
    "131 +/- 18.8", "74.8 +/- 9.4",
    as.character(round(auc600$auc[auc600$R == 0], 1)),
    as.character(round(auc600$auc[auc600$R == 10], 1))
  ),
  `CLR (mL/min)` = c("0.415 +/- 0.06", "0.910 +/- 0.10", "0.540 (published)", "1.072 (published)"),
  `Sleep time (min)` = c("126 +/- 15.1", "87.0 +/- 11.1", "not modelled", "not modelled")
) |>
  knitr::kable(caption = "Replicates Table III of Felmlee 2010 alongside the corresponding simulated arms.")
```

```{r table3-assert}
# The observed GHB-alone AUC (131 min*mg/mL, N = 3) is an independent check on
# the 600 mg/kg typical-value solve: it comes from a different experiment than
# the one that produced Table II.
obs_alone <- 131
stopifnot(abs(auc600$auc[auc600$R == 0] - obs_alone) / obs_alone < 0.15)
```

The typical-value solve at 600 mg/kg lands within a few percent of the
independently observed AUC of 131 min*mg/mL. The observed L-lactate arm (74.8)
falls below the simulated `R = 10` AUC (96.1 published); the paper attributes
this to L-lactate additionally altering tissue distribution, an effect the model
deliberately does not include ("we assumed that the hypothetical inhibitor would
alter active renal reabsorption and not influence tissue distribution or
metabolism of GHB").

The sedative/hypnotic sleep-time endpoint of Table III is a pharmacodynamic
measurement with no fitted model in this paper, so it is reported here for
context but is not part of the packaged model.

# Assumptions and deviations

## Errata and source ambiguities

**GFR units are misprinted in Table I.** Table I gives GFR as `10` with the unit
`mg/min/kg`. That is dimensionally impossible in Eqs. 5-6, where GFR multiplies
a concentration to produce a mass rate, so it must be a volumetric flow;
`mg` is a typo for `mL`. Two readings remained, and simulating both against the
paper's own Table II "GHB alone" column decides between them. The comparison is
computed here rather than asserted, so the numbers and the conclusion cannot
drift apart:

```{r gfr-reading}
# The packaged model computes gfr <- exp(lgfr) * WT, so the per-kg reading is
# lgfr = log(10) (giving 3.0 mL/min at 0.3 kg) and the absolute reading is
# lgfr = log(10 / WT) (giving 10 mL/min regardless of weight).
gfr_auc <- function(lgfr_val) {
  vapply(DOSES, function(d) {
    s <- rxode2::rxSolve(
      mod, build_events(d, tgrid = seq(0, 360, by = 0.5)),
      omega = NA, sigma = NA, addDosing = FALSE,
      params = c(lgfr = lgfr_val), returnType = "data.frame"
    )
    s <- s[!is.na(s$Cc), ]
    trap_auc(s$time, s$Cc)
  }, numeric(1))
}

gfr_cmp <- data.frame(
  dose      = DOSES,
  published = c(29.3, 83.0, 133.3, 208.2),
  per_kg    = gfr_auc(log(10)),
  absolute  = gfr_auc(log(10 / WT_KG))
) |>
  dplyr::mutate(
    per_kg_pct   = 100 * (per_kg - published) / published,
    absolute_pct = 100 * (absolute - published) / published
  )
```

```{r gfr-reading-table, echo = FALSE}
gfr_cmp |>
  dplyr::transmute(
    `GHB dose (mg/kg)`        = dose,
    `Published AUC`           = published,
    `GFR 10 mL/min/kg`        = round(per_kg, 1),
    `% diff`                  = round(per_kg_pct, 1),
    `GFR 10 mL/min absolute`  = round(absolute, 1),
    `% diff `                 = round(absolute_pct, 1)
  ) |>
  knitr::kable(
    caption = "Which GFR unit reading reproduces Table II. Deterministic typical-value solves."
  )
```

```{r gfr-reading-assert}
# Both sides are typical-value solves with no random draws, so exact bounds are
# reproducible across rxode2 builds. The per-kg reading tracks the published
# column; the absolute reading is low by a wide margin at every dose.
stopifnot(
  max(abs(gfr_cmp$per_kg_pct)) < 10,
  all(gfr_cmp$absolute_pct < -30),
  # And it is decisively the better reading dose by dose, not just on average.
  all(abs(gfr_cmp$per_kg_pct) < abs(gfr_cmp$absolute_pct))
)
```

The per-kilogram reading is therefore correct: it lands within 10% of the
published AUC at every dose, whereas the absolute reading is more than 30% low
at every dose (worst at the highest doses, where it misses by roughly a factor
of three). That makes body weight the only covariate in the model
(`gfr <- exp(lgfr) * WT`). Every other Table I flow and volume is an absolute
per-rat value.

**GFR is encoded as `fixed()` although Table I does not mark it.** Table I's
footnote a ("Parameter was fixed to physiological value") appears on Vplasma,
Vkidney, Vulf1, Vulf2, QR and UF, but not on GFR. The packaged model
nonetheless wraps GFR in `fixed()`, on three grounds: it is the only row whose
units are per kilogram (the form physiology compendia use, not the form a NONMEM
THETA would take in a model where nothing else is weight-scaled); it is the only
non-error row printed as a bare integer while every estimated parameter carries
three significant figures; and the Results state the model fixed parameters to
physiological values citing references 27-29, one of which is Davies and Morris
1993, the standard laboratory-animal physiology compendium. The distinction is
provenance-only: the value and every simulation in this vignette are identical
either way. A reader who prefers the literal footnote reading should treat
`lgfr` as estimated.

**Body weight is not reported per animal.** The paper gives a range of
280-320 g. This vignette uses 0.3 kg. Independent corroboration: Table I's fixed
QR of 12.5 mL/min equals the rat kidney blood flow implied by standard
physiology (14.1% of a cardiac output of 296 mL/min/kg) at almost exactly
0.3 kg.

**Table II's CL_R column is not fully reconstructed.** Eq. 14 defines
`CLR = Ae,inf / AUC_plasma`. Cumulative urinary excretion is complete by 360 min
in this model (extending the horizon to 5,760 min does not change `Ae`), so the
numerator is unambiguous. Reproducing the published CL_R values nevertheless
requires the population-mean quantities described in the Figure 2 section above,
and even then the simulated values fall short at the lower doses, most at
200 mg/kg, converging on the published ones as dose rises. The per-dose
shortfall is not itself stable across random draws, because it is the mean of a
heavy-tailed distribution; the pattern of "worst at the lowest dose" is.
The plasma AUC column, which is the better-determined
quantity and the one the model was fitted to, reproduces throughout. The
residual gap sits in a derived metric that depends on the mean of a
strongly right-skewed distribution generated by a 114% CV parameter, and on
reconstruction choices the paper does not fully specify (the exact per-animal
weights, and whether the reported means were taken over simulated observations
or individual predictions). No parameter was adjusted to close it.

**Number of animals.** The paper reports 7-10 rats per dose group across four
groups without exact per-group counts; `population$n_subjects` records 34, the
midpoint of the implied 28-40 range.

## Modelling choices

* **Residual error.** Eqs. 9-10 specify `Y = Log(prediction) + epsilon`, i.e.
  additive on the natural-log scale, which is nlmixr2's `lnorm()` residual-error
  model, not `prop()`.
* **Between-subject variability scale.** Table I reports CV%, and the paper
  states it followed the variability reporting of Fung et al. (reference 26).
  The packaged model uses `omega^2 = log(1 + CV^2)`, matching the sibling model
  `Fung_2008_butanediol_rat.R` already in the library. The alternative reading
  (`omega^2 = CV^2`) was tested against the mean and the 10th / 90th percentiles
  of Figure 2 at all four doses and is not distinguishable from it on that
  evidence, so the library-consistent convention was kept.
* **Inhibition extension.** Encoded with two multiplier covariates rather than a
  single ratio plus a mechanism switch, so that all three published mechanisms
  are expressed in one branch-free rate law that reduces exactly to the fitted
  Eq. 4 when both covariates are 0.
* **Cohort size.** This vignette simulates `r NSUB` rats per dose where the
  paper simulated 1,000, to keep the vignette within the package's render-time
  budget.
* **Not modelled.** The sedative/hypnotic sleep-time endpoint of Table III, and
  any effect of L-lactate on tissue distribution.

# Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
