---
title: "Imipenem (Feng 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Imipenem (Feng 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Feng_2026_imipenem_pbpk"))$reference`
* Article: <https://doi.org/10.3389/fcimb.2026.1798911>
* Supplement: <https://www.frontiersin.org/articles/10.3389/fcimb.2026.1798911/full#supplementary-material>

```{r description, echo = FALSE, results = "asis"}
cat(rxode2::rxode(readModelDb("Feng_2026_imipenem_pbpk"))$description)
```

Feng 2026 built a whole-body physiologically based pharmacokinetic (PBPK) model
of imipenem in GastroPlus 9.9, verified it against eight published clinical
studies in healthy adults, adults with renal impairment (RI) and children, and
then extrapolated it to children aged 3-18 years with RI in order to propose
dose adjustments. The GastroPlus whole-body structure -- its ODEs, organ
volumes, blood flows and tissue-to-plasma partition coefficients -- is not
written out anywhere in the paper, so it is not reproduced here.

What the paper *does* publish, in Supplementary Table S3, is a complete reduced
disposition card: total clearance, steady-state volume of distribution,
half-life, glomerular filtration rate and unbound fraction for every population
it simulated. That card is what this model file encodes, following the
`Yang_2024_meropenem_pbpk` precedent for a platform-PBPK paper that carries a
reduced model into its own dosing analysis.

# Population

Eight clinical studies were pooled from the literature and digitised with
GetData Graph Digitizer 2.26. Six are adult (Supplementary Table S1): Wang 2021
(China, n = 12, 34.2 years, 60.4 kg), Jaruratanasirikul 2005 (Thailand, n = 8,
28.3 years, 58.8 kg), Nilsson 1991 (Sweden, n = 8, 33 years, 74 kg), Norrby 1983
(Sweden, n = 8, 25 years, 75 kg), Drusano 1984 (USA, n = 6, 19-34 years) and
Gibson 1985 (USA, n = 6 each in mild, moderate and severe RI, 47.7-60.8 years,
71.3-86.6 kg). Two are paediatric (Supplementary Table S2): Claesson 1992
(Sweden, children 3-12 years with peritonitis, n = 8 at 15 mg/kg and n = 2 at
25 mg/kg) and Bradley 2023 (USA, children 2-6 years with gram-negative
infections, n = 6, 15.4 kg).

Every study administered imipenem together with cilastatin or another enzyme
inhibitor, so the renal clearance the model carries is the cilastatin-inhibited
value (Table 1 footnote). Renal function is classified per FDA guidance on
glomerular filtration rate: normal >= 90, mild 60-89, moderate 30-59 and severe
15-29 mL/min/1.73 m^2. The paediatric age bands are the WHO ones the paper
adopts: preschool 3-6, school-age 6-12 and adolescent 12-18 years, represented
in the simulations by 3-, 8- and 16-year-old children.

**There are no paediatric renal-impairment PK data anywhere in the source.** The
paper says so explicitly in its Limitations, and the paediatric RI arm is a pure
extrapolation of the software's built-in renal physiology.

```{r population, echo = FALSE}
str(readModelDb("Feng_2026_imipenem_pbpk")()$population)
```

# Source trace

Per-parameter origins are recorded as in-file comments beside each `ini()` entry
in `inst/modeldb/specificDrugs/Feng_2026_imipenem_pbpk.R`; they are collected
here for review.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lvcPerKg` | 0.1636 L/kg | Supplementary Table S3 `Vss` divided by the Supplementary Table S1 weights (seven rows, 0.1555-0.1730) |
| `lcl_adult` | 11.176 L/h | Supplementary Table S3, Wang 2021 q6h / Drusano 1984 rows; Discussion quotes "approximately 11.126 L/h" |
| `e_renalimp_mild_cl_adult` | 0.95580 | Supplementary Table S3, Gibson mild `CL` 10.682 / 11.176 |
| `e_renalimp_mod_cl_adult` | 0.57051 | Supplementary Table S3, Gibson moderate `CL` 6.376 / 11.176 |
| `e_renalimp_sev_cl_adult` | 0.31192 | Supplementary Table S3, Gibson severe `CL` 3.486 / 11.176 |
| `lclPerKg_preschool` | 0.47835 L/h/kg | Table 4 preschool normal, 15 / 31.358 |
| `lclPerKg_schoolage` | 0.35032 L/h/kg | Table 4 school-age normal, 15 / 42.818 |
| `lclPerKg_adolescent` | 0.17280 L/h/kg | Table 4 adolescent normal, 15 / 86.804 |
| `e_renalimp_*_cl_preschool` | 0.96246 / 0.80554 / 0.49650 | Table 4 preschool, 31.358 / (32.581, 38.928, 63.158) |
| `e_renalimp_*_cl_schoolage` | 0.95723 / 0.79639 / 0.46826 | Table 4 school-age, 42.818 / (44.731, 53.765, 91.441) |
| `e_renalimp_*_cl_adolescent` | 0.94831 / 0.74580 / 0.42434 | Table 4 adolescent, 86.804 / (91.535, 116.39, 204.56) |
| `fup_adult` | 0.80000 | Table 1 `Fup` 80%; Supplementary Table S3 healthy adult rows |
| `fup_adult_mild` / `_mod` / `_sev` | 0.80588 / 0.82723 / 0.83352 | Supplementary Table S3, Gibson rows |
| `fup_ped` | 0.81217 | Supplementary Table S3, Claesson 1992 row (Bradley 2023 row gives 0.81182) |
| `addSd`, `propSd` | `fixed(0)` | Not reported anywhere in the paper or its supplement |
| WHO age bands 3 / 6 / 12 / 18 years | n/a | Methods, "Prediction of imipenem exposures in pediatric patients with RI" |
| FDA renal classification | n/a | Methods, "Model for adult patients with RI" |
| `d/dt(central) <- -kel * central` | n/a | Reduction of the GastroPlus whole-body model; justified below |
| 70%fT>MIC target, PTA >= 80% | n/a | Methods, "Pharmacodynamic evaluation of imipenem" |

# Why a one-compartment reduction is the right shape

Supplementary Table S3 reports `CL`, `Vss` and `T1/2` for every simulated
population -- 15 rows, of which 13 carry distinct parameter triples (the two
Jaruratanasirikul 0.5 g rows are identical to each other, as are the two
Claesson rows, so the table below lists each triple once). If those three
numbers describe a mono-exponential disposition then `T1/2 = ln2 * Vss / CL`
must hold exactly. It does, on every row.

The same check settles a unit error. Supplementary Table S3 heads its volume
column `Vss(L/kg)`, and the Results text repeats it ("about 3.259 L/kg in
children, compared with 11.663 L/kg in adults"). That cannot be right: 11.663
L/kg in a 74 kg adult is 863 L, which with the tabulated 11.126 L/h clearance
gives a 54 h half-life for a drug the same row calls 0.723 h. The identity below
closes only when `Vss` is read as absolute litres.

```{r s3-identity}
s3 <- tibble::tribble(
  ~study,                 ~renal,        ~CL,     ~Vss,   ~thalf,  ~WT,
  "Wang 2021 0.25 g",     "Normal",   10.972,  10.447,   0.660,  60.40,
  "Jaruratanasirikul 0.5 g", "Normal", 10.726,  9.135,   0.590,  58.75,
  "Jaruratanasirikul 1.0 g", "Normal", 10.726,  9.135,   0.590,  58.75,
  "Nilsson 1991 1.0 g",   "Normal",   11.126,  11.663,   0.723,  74.00,
  "Norrby 1983 0.5 g",    "Normal",   11.262,  12.371,   0.761,  75.00,
  "Norrby 1983 1.0 g",    "Normal",   11.262,  12.371,   0.761,  75.00,
  "Wang 2021 q6h",        "Normal",   11.176,  11.663,   0.723,  NA_real_,
  "Drusano 1984 q6h",     "Normal",   11.176,  11.663,   0.723,  NA_real_,
  "Gibson 1985",          "Mild RI",  10.682,  14.081,   0.913,  86.60,
  "Gibson 1985",          "Moderate RI", 6.376, 12.275,   1.334,  76.50,
  "Gibson 1985",          "Severe RI",   3.486, 12.194,   2.424,  71.30,
  "Claesson 1992",        "Normal",    7.886,   3.259,   0.286,  NA_real_,
  "Bradley 2023",         "Normal",    8.489,   4.221,   0.345,  NA_real_
) |>
  mutate(
    thalf_identity = log(2) * Vss / CL,
    pct_error      = 100 * (thalf_identity - thalf) / thalf,
    VssPerKg       = Vss / WT
  )

s3 |>
  select(study, renal, CL, Vss, thalf, thalf_identity, pct_error, VssPerKg) |>
  rename(
    "Study" = study, "Renal function" = renal,
    "CL (L/h)" = CL, "Vss (L)" = Vss,
    "T1/2 published (h)" = thalf, "ln2*Vss/CL (h)" = thalf_identity,
    "Error (%)" = pct_error, "Vss/WT (L/kg)" = VssPerKg
  ) |>
  knitr::kable(
    digits = c(0, 0, 3, 3, 3, 4, 3, 4),
    caption = paste(
      "Supplementary Table S3 of Feng 2026 is internally mono-exponential.",
      "The published half-life is recovered from ln2 * Vss / CL on every row,",
      "which is only dimensionally possible with Vss in absolute litres --",
      "falsifying the printed 'L/kg' unit. Vss / body weight is then a",
      "physiological 0.155-0.173 L/kg across 58.75-86.6 kg adults."
    )
  )

# Deterministic arithmetic on published numbers, so a tight bound is correct
# here; this is not a cohort-derived quantity.
stopifnot(max(abs(s3$pct_error)) < 0.6)
stopifnot(all(s3$VssPerKg > 0.15 & s3$VssPerKg < 0.18, na.rm = TRUE))
```

The volume the model carries, 0.1636 L/kg, is the mean of the seven `Vss / WT`
values above that have a published weight. It is confirmed independently below
by an 8-year-old child.

# Virtual cohort

The packaged model has no interindividual variability and no residual error --
Feng 2026 publishes neither -- so it is fully deterministic and one subject per
scenario is sufficient. Every arm below is far under the 200-per-arm cap.

Two arms are simulated: the ten adult regimens of Table 2 (each at its own
study's body weight from Supplementary Table S1) and the paediatric regimens of
Tables 3 and 4.

```{r cohort}
# One subject per published regimen. `tend` is set to at least a dozen
# half-lives so the terminal phase is fully characterised for NCA, without
# running so far that concentrations underflow and PKNCA's log-down
# trapezoid takes the log of a round-off negative.
make_subject <- function(id, label, arm, amt, dur, WT, AGE,
                         mild = 0, mod = 0, sev = 0, tend = 24) {
  # Round before de-duplicating: the two seq() grids generate the same nominal
  # time (e.g. 0.15) by different arithmetic, so without rounding `unique()`
  # leaves near-duplicate times that differ in the last bit.
  grid <- sort(unique(round(c(
    seq(0, min(4, tend), by = 0.005),
    seq(0, tend, by = 0.05)
  ), 6)))
  dplyr::bind_rows(
    tibble::tibble(id = id, time = 0, evid = 1L, amt = amt, dur = dur,
                   cmt = "central"),
    tibble::tibble(id = id, time = grid, evid = 0L, amt = NA_real_, dur = 0,
                   cmt = "central")
  ) |>
    dplyr::mutate(
      label = label, arm = arm, WT = WT, AGE = AGE,
      RENALIMP_MILD = mild, RENALIMP_MOD = mod, RENALIMP_SEV = sev
    )
}

# ---- Arm 1: adults, Table 2 (weights from Supplementary Table S1) ----------
adult_spec <- tibble::tribble(
  ~label,                        ~amt, ~dur,  ~WT,  ~mild, ~mod, ~sev, ~tend,
  "Wang 0.25 g / 0.5 h",          250, 0.50, 60.40,     0,    0,    0,     10,
  "Jarurat. 0.5 g / 0.5 h",       500, 0.50, 58.75,     0,    0,    0,     10,
  "Nilsson 1.0 g / 0.5 h",       1000, 0.50, 74.00,     0,    0,    0,     10,
  "Jarurat. 0.5 g / 2 h",         500, 2.00, 58.75,     0,    0,    0,     12,
  "Jarurat. 1.0 g / 2 h",        1000, 2.00, 58.75,     0,    0,    0,     12,
  "Norrby 0.5 g bolus",           500, 0.00, 75.00,     0,    0,    0,     10,
  "Norrby 1.0 g bolus",          1000, 0.00, 75.00,     0,    0,    0,     10,
  "Gibson mild 0.25 g",           250, 0.15, 86.60,     1,    0,    0,     12,
  "Gibson moderate 0.25 g",       250, 0.15, 76.50,     0,    1,    0,     18,
  "Gibson severe 0.25 g",         250, 0.15, 71.30,     0,    0,    1,     30
)
```

```{r cohort-peds}
# ---- Arm 2: children, Tables 3 and 4 --------------------------------------
# A mg/kg dose makes both AUC and Cmax independent of body weight in this
# model (dose and clearance both scale with WT, and so does volume), so the
# representative weights below only have to be plausible -- they cancel.
# `tend` is set to roughly 13-20 half-lives in every row. Longer is NOT better:
# the paediatric half-lives are as short as 0.24 h, so a 24 h window would run
# the amount in the central compartment down past the solver's absolute
# tolerance, and PKNCA's log-down trapezoid then takes the log of a round-off
# negative (known-vignette-failure-patterns: log-down NaNs on negative
# round-off). Thirteen half-lives already leaves < 0.02% of the dose.
ped_spec <- tibble::tribble(
  ~label,                          ~AGE, ~WT, ~mgkg, ~dur, ~mild, ~mod, ~sev, ~tend,
  "Preschool 3 y, normal",            3,  14,    15, 0.50,     0,    0,    0,     5,
  "Preschool 3 y, mild RI",           3,  14,    15, 0.50,     1,    0,    0,     5,
  "Preschool 3 y, moderate RI",       3,  14,    15, 0.50,     0,    1,    0,     6,
  "Preschool 3 y, severe RI",         3,  14,    15, 0.50,     0,    0,    1,     8,
  "School-age 8 y, normal",           8,  25,    15, 0.50,     0,    0,    0,     6,
  "School-age 8 y, mild RI",          8,  25,    15, 0.50,     1,    0,    0,     6,
  "School-age 8 y, moderate RI",      8,  25,    15, 0.50,     0,    1,    0,     8,
  "School-age 8 y, severe RI",        8,  25,    15, 0.50,     0,    0,    1,    12,
  "Adolescent 16 y, normal",         16,  55,    15, 0.50,     0,    0,    0,    12,
  "Adolescent 16 y, mild RI",        16,  55,    15, 0.50,     1,    0,    0,    12,
  "Adolescent 16 y, moderate RI",    16,  55,    15, 0.50,     0,    1,    0,    16,
  "Adolescent 16 y, severe RI",      16,  55,    15, 0.50,     0,    0,    1,    26
)
```

```{r build-events}
build_arm <- function(spec, arm, id_offset, dose_fn, age_fn = NULL) {
  out <- vector("list", nrow(spec))
  for (i in seq_len(nrow(spec))) {
    r <- spec[i, ]
    out[[i]] <- make_subject(
      id    = id_offset + i,
      label = r$label,
      arm   = arm,
      amt   = dose_fn(r),
      dur   = r$dur,
      WT    = r$WT,
      AGE   = if (is.null(age_fn)) 35 else r$AGE,
      mild  = r$mild, mod = r$mod, sev = r$sev,
      tend  = r$tend
    )
  }
  dplyr::bind_rows(out)
}

events <- dplyr::bind_rows(
  build_arm(adult_spec, "Adult",     0L, function(r) r$amt),
  build_arm(ped_spec,   "Paediatric", 100L, function(r) r$mgkg * r$WT, age_fn = TRUE)
)

# Disjoint ids across arms (a duplicate id silently merges two subjects).
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
stopifnot(length(unique(events$id)) == nrow(adult_spec) + nrow(ped_spec))
```

# Simulation

```{r simulate}
mod <- readModelDb("Feng_2026_imipenem_pbpk")

# No etas are declared, so `omega = NA` / `zeroRe()` would error
# (known-vignette-failure-patterns.md pattern 9) and is unnecessary: the
# model is already deterministic.
sim <- rxode2::rxSolve(
  mod, events = events,
  keep = c("label", "arm", "WT", "AGE",
           "RENALIMP_MILD", "RENALIMP_MOD", "RENALIMP_SEV")
) |>
  as.data.frame() |>
  # `keep` can return character columns as factors; downstream joins are on
  # `label` against character literals, which would error on a factor.
  dplyr::mutate(label = as.character(label), arm = as.character(arm))

stopifnot(nrow(sim) > 0, !anyNA(sim$Cc))
# Free concentration must be the unbound fraction of total wherever there is
# drug. The ratio is tested only where Cc > 0 -- at t = 0 on an infusion row
# both are exactly 0 and a strict ratio bound is undefined, not violated.
pos <- sim$Cc > 0
stopifnot(any(pos))
fup_ratio <- sim$Cfree[pos] / sim$Cc[pos]
# Every fup the model can resolve lies in [0.800, 0.834] (Supplementary Table
# S3), so this pins the covariate wiring, not just the sign.
stopifnot(all(fup_ratio >= 0.7999), all(fup_ratio <= 0.8340))
```

The individual parameters the model resolves for each stratum are worth showing
directly, because they are the whole content of the reduction.

```{r params}
sim |>
  group_by(arm, label) |>
  summarise(CL = first(cl), Vc = first(vc), fup = first(fup), .groups = "drop") |>
  mutate(`T1/2 (h)` = log(2) * Vc / CL) |>
  rename("Arm" = arm, "Regimen" = label, "CL (L/h)" = CL, "Vc (L)" = Vc,
         "Unbound fraction" = fup) |>
  knitr::kable(digits = 4, caption = "Model-resolved disposition per simulated regimen.")
```

# Replicate Figure 5

Figure 5 of Feng 2026 shows the predicted plasma concentration-time profile of
imipenem in a representative school-age child after 15 mg/kg, panel A as a
0.5 h infusion and panel B as a 3 h infusion.

```{r figure-5}
fig5_spec <- tidyr::crossing(
  tibble::tibble(renal = c("Normal", "Mild RI", "Moderate RI", "Severe RI"),
                 mild  = c(0, 1, 0, 0), mod = c(0, 0, 1, 0), sev = c(0, 0, 0, 1)),
  tibble::tibble(infusion = c("A: 0.5 h infusion", "B: 3 h infusion"),
                 dur = c(0.5, 3.0))
)

fig5_events <- dplyr::bind_rows(lapply(seq_len(nrow(fig5_spec)), function(i) {
  r <- fig5_spec[i, ]
  make_subject(id = 500L + i, label = r$renal, arm = r$infusion,
               amt = 15 * 25, dur = r$dur, WT = 25, AGE = 8,
               mild = r$mild, mod = r$mod, sev = r$sev, tend = 6)
}))

fig5 <- rxode2::rxSolve(mod, events = fig5_events,
                        keep = c("label", "arm")) |>
  as.data.frame() |>
  dplyr::mutate(label = as.character(label), arm = as.character(arm))

ggplot(fig5, aes(time, Cc, colour = label)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~arm) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Imipenem plasma concentration (mg/L)",
       colour = "Renal function",
       title = "Figure 5 -- school-age child (8 years), 15 mg/kg",
       caption = "Replicates Figure 5 of Feng 2026.")
```

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, label, arm)

# Guarantee a time = 0 row per subject. Imipenem is given intravenously with
# no prior exposure, so the pre-dose concentration is 0.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, label, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | label + id,
                             concu = "mg/L", timeu = "h")

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, label, arm)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | label + id, doseu = "mg")

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
```

## Comparison against the paper's own PBPK predictions

The reduction's job is to reproduce what the whole-body GastroPlus model
predicted, so the paper's *predicted* AUC and Cmax are the reference. Adult
values are Table 2 and paediatric values are Tables 3 and 4. The paper reports
AUC0-t (truncated at each study's last sampling time) whereas the simulation
reports AUC0-inf, so the simulated AUC should sit at or slightly above the
published one.

```{r nca-compare}
published <- tibble::tribble(
  ~label,                          ~cmax,  ~aucinf.obs,
  "Wang 0.25 g / 0.5 h",           19.21,        22.59,
  "Jarurat. 0.5 g / 0.5 h",        43.62,        50.91,
  "Nilsson 1.0 g / 0.5 h",         71.15,        88.47,
  "Jarurat. 0.5 g / 2 h",          20.56,        46.20,
  "Jarurat. 1.0 g / 2 h",          41.12,        92.39,
  "Norrby 0.5 g bolus",           165.99,        43.86,
  "Norrby 1.0 g bolus",           331.99,        87.72,
  "Gibson mild 0.25 g",            23.03,        23.13,
  "Gibson moderate 0.25 g",        25.67,        37.69,
  "Gibson severe 0.25 g",          25.87,        66.65,
  "Preschool 3 y, normal",            NA,       31.358,
  "Preschool 3 y, mild RI",           NA,       32.581,
  "Preschool 3 y, moderate RI",       NA,       38.928,
  "Preschool 3 y, severe RI",         NA,       63.158,
  "School-age 8 y, normal",        56.42,       42.818,
  "School-age 8 y, mild RI",       57.37,       44.731,
  "School-age 8 y, moderate RI",   61.14,       53.765,
  "School-age 8 y, severe RI",     71.58,       91.441,
  "Adolescent 16 y, normal",          NA,       86.804,
  "Adolescent 16 y, mild RI",         NA,       91.535,
  "Adolescent 16 y, moderate RI",     NA,      116.390,
  "Adolescent 16 y, severe RI",       NA,      204.560
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "label",
  units         = c(cmax = "mg/L", aucinf.obs = "mg*h/L"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Reduced model vs the whole-body GastroPlus predictions of Feng 2026.",
    "* marks a difference above 20%. Every starred cell is a Cmax, and every",
    "one of them is a regimen whose input is faster than the distribution the",
    "reduction cannot see: the two instantaneous boluses and the two shortest",
    "(0.15 h) infusions. No AUC is starred, and no Cmax on any regimen the",
    "paper actually recommends is starred. Discussed below."
  )
)
```

The comparison is quantified below so it can be asserted rather than eyeballed.
Every quantity here is deterministic -- the model has no random effects -- so
tight bounds are appropriate and will catch a regression.

```{r nca-scorecard}
sim_wide <- as.data.frame(nca_res) |>
  dplyr::select(label, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

score <- published |>
  dplyr::left_join(sim_wide, by = "label", suffix = c("_pub", "_sim")) |>
  dplyr::mutate(
    infusion    = !grepl("bolus", label),
    auc_ratio   = aucinf.obs_sim / aucinf.obs_pub,
    cmax_ratio  = cmax_sim / cmax_pub
  )

# The join must have matched every published row (a silent zero-row lookup
# would make every assertion below vacuously true).
stopifnot(nrow(score) == nrow(published), !anyNA(score$auc_ratio))

score |>
  dplyr::select(label, aucinf.obs_pub, aucinf.obs_sim, auc_ratio,
                cmax_pub, cmax_sim, cmax_ratio) |>
  dplyr::rename(
    "Regimen" = label,
    "AUC published" = aucinf.obs_pub, "AUC simulated" = aucinf.obs_sim,
    "AUC ratio" = auc_ratio,
    "Cmax published" = cmax_pub, "Cmax simulated" = cmax_sim,
    "Cmax ratio" = cmax_ratio
  ) |>
  knitr::kable(digits = 3,
               caption = "Reduced model / published PBPK prediction, per regimen.")

is_ped <- grepl("^(Preschool|School-age|Adolescent)", score$label)

# Paediatric AUC is an exact identity, and that is the point of asserting it:
# each band's CL/kg was transcribed as 15 / AUC0-24h, and both dose and
# clearance scale with body weight, so the reduction must return all twelve
# Table 4 cells to numerical precision. This checks the transcription of the
# three band clearances, the nine renal multipliers and the weight
# cancellation simultaneously. A single mistyped digit breaks it.
stopifnot(sum(is_ped) == 12L)
stopifnot(max(abs(score$auc_ratio[is_ped] - 1)) < 0.005)

# Adult AUC. One regimen is an outlier -- Jaruratanasirikul 0.5 g / 0.5 h, at
# 0.879 -- because the paper's own printed AUC for that row exceeds what its
# own tabulated clearance allows (see Assumptions and deviations). Every other
# adult regimen is within 5%.
adult_auc <- score$auc_ratio[!is_ped]
stopifnot(length(adult_auc) == 10L)
stopifnot(all(adult_auc > 0.87 & adult_auc < 1.08))
stopifnot(sum(abs(adult_auc - 1) < 0.05) == 8L)

# Cmax is reproduced on every INFUSION regimen. The two instantaneous-bolus
# rows are excluded and discussed in Assumptions and deviations.
cmax_inf <- score$cmax_ratio[score$infusion & !is.na(score$cmax_ratio)]
stopifnot(length(cmax_inf) == 12L)
stopifnot(all(cmax_inf > 0.70 & cmax_inf < 1.05))

# The four school-age rows -- the subjects the whole dose recommendation rests
# on -- are reproduced far more tightly than that.
schoolage <- score$cmax_ratio[grepl("School-age", score$label)]
stopifnot(length(schoolage) == 4L, all(abs(schoolage - 1) < 0.02))
```

## Comparison against the clinical observations

A separate question from "does the reduction match the platform" is "does it
match the patients". Feng 2026 reports observed AUC0-t for eight of the adult
regimens and three paediatric ones (Tables 2 and 3).

```{r observed-compare}
observed <- tibble::tribble(
  ~label,                      ~auc_obs, ~cmax_obs,
  "Wang 0.25 g / 0.5 h",          25.48,     20.05,
  "Jarurat. 0.5 g / 0.5 h",       60.65,     48.38,
  "Nilsson 1.0 g / 0.5 h",        89.37,     65.64,
  "Jarurat. 0.5 g / 2 h",         57.20,     21.23,
  "Jarurat. 1.0 g / 2 h",        116.81,     42.33,
  "Norrby 0.5 g bolus",           42.35,        NA,
  "Norrby 1.0 g bolus",           95.75,        NA,
  "Gibson mild 0.25 g",           19.97,        NA,
  "Gibson moderate 0.25 g",       29.80,        NA,
  "Gibson severe 0.25 g",         60.76,        NA
)

obs_cmp <- observed |>
  dplyr::left_join(sim_wide, by = "label") |>
  dplyr::mutate(auc_fold = aucinf.obs / auc_obs,
                cmax_fold = cmax / cmax_obs)

stopifnot(nrow(obs_cmp) == nrow(observed), !anyNA(obs_cmp$auc_fold))

obs_cmp |>
  dplyr::select(label, auc_obs, aucinf.obs, auc_fold, cmax_obs, cmax, cmax_fold) |>
  dplyr::rename("Regimen" = label,
                "AUC observed" = auc_obs, "AUC simulated" = aucinf.obs,
                "AUC fold-error" = auc_fold,
                "Cmax observed" = cmax_obs, "Cmax simulated" = cmax,
                "Cmax fold-error" = cmax_fold) |>
  knitr::kable(digits = 3,
               caption = paste(
                 "Reduced model vs the clinical observations Feng 2026",
                 "digitised. The paper's own acceptance criterion is a",
                 "fold error below 2."))

# The paper's own acceptance criterion, applied to the reduction.
stopifnot(all(obs_cmp$auc_fold > 0.5 & obs_cmp$auc_fold < 2))
cmax_fold <- obs_cmp$cmax_fold[!is.na(obs_cmp$cmax_fold)]
stopifnot(length(cmax_fold) == 5L, all(cmax_fold > 0.5 & cmax_fold < 2))
```

# Do the proposed dose adjustments equalise exposure?

This is the paper's headline result. Feng 2026 derives the recommended dose by
multiplying the standard 15 mg/kg by the reciprocal of the RI-to-healthy
geometric mean ratio, arriving at 15, 15, 12 and 7 mg/kg every 6 hours for
normal renal function, mild, moderate and severe RI. If the reduction is
faithful, those four doses should all deliver the exposure a healthy child gets
from 15 mg/kg.

```{r dose-adjust}
dose_spec <- tibble::tribble(
  ~renal,          ~mgkg, ~mild, ~mod, ~sev, ~tend,
  "Normal",           15,     0,    0,    0,     8,
  "Mild RI",          15,     1,    0,    0,     8,
  "Moderate RI",      12,     0,    1,    0,    10,
  "Severe RI",         7,     0,    0,    1,    16
)

dose_events <- dplyr::bind_rows(lapply(seq_len(nrow(dose_spec)), function(i) {
  r <- dose_spec[i, ]
  make_subject(id = 700L + i, label = r$renal, arm = "Proposed dose",
               amt = r$mgkg * 25, dur = 3.0, WT = 25, AGE = 8,
               mild = r$mild, mod = r$mod, sev = r$sev, tend = r$tend)
}))

dose_sim <- rxode2::rxSolve(mod, events = dose_events,
                            keep = c("label")) |>
  as.data.frame() |>
  dplyr::mutate(label = as.character(label))

dose_given <- dose_events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(label, amt)

dose_auc <- dose_sim |>
  dplyr::group_by(label) |>
  dplyr::summarise(CL = first(cl), .groups = "drop") |>
  dplyr::left_join(dose_given, by = "label") |>
  # AUC0-inf = dose / CL exactly for this linear one-compartment model.
  dplyr::mutate(auc = amt / CL) |>
  dplyr::arrange(match(label, dose_spec$renal))

stopifnot(nrow(dose_auc) == nrow(dose_spec), !anyNA(dose_auc$auc))

ref_auc <- dose_auc$auc[dose_auc$label == "Normal"]
stopifnot(length(ref_auc) == 1L)

dose_auc |>
  dplyr::mutate(ratio_to_normal = auc / ref_auc) |>
  dplyr::rename("Renal function" = label, "AUC0-inf (mg*h/L)" = auc,
                "CL (L/h)" = CL, "Dose (mg)" = amt,
                "Ratio to normal-renal child" = ratio_to_normal) |>
  knitr::kable(digits = 3,
               caption = paste(
                 "Exposure delivered by the doses Feng 2026 proposes",
                 "(Table 4), in an 8-year-old, 25 kg child."))

# All four proposed doses land within 5% of the healthy-child exposure.
stopifnot(all(abs(dose_auc$auc / ref_auc - 1) < 0.05))
```

# Prolonged infusion and %fT>MIC

Imipenem is time-dependent, and the paper's PK/PD index is the percentage of the
dosing interval during which the *free* concentration exceeds the MIC, with
70%fT>MIC as the target. Feng 2026's second headline result is that extending
the infusion from 0.5 h to 3 h substantially raises target attainment.

Probability of target attainment itself cannot be reproduced here: it is a
population quantity and Feng 2026 publishes no interindividual variance for any
parameter. What *is* reproducible is the deterministic %fT>MIC for the typical
child, and the direction and size of the infusion-duration effect.

```{r ftmic}
ftmic_spec <- tidyr::crossing(
  tibble::tibble(renal = c("Normal", "Mild RI", "Moderate RI", "Severe RI"),
                 mild = c(0, 1, 0, 0), mod = c(0, 0, 1, 0), sev = c(0, 0, 0, 1)),
  tibble::tibble(inf_label = c("0.5 h", "3 h"), dur = c(0.5, 3.0))
)

# q6h. The model's paediatric half-lives are 0.32-0.69 h, i.e. at least 8.7
# dosing-interval half-lives, so accumulation is nil and the first interval
# after a single dose is the steady-state interval to within rounding.
tau <- 6

ftmic_events <- dplyr::bind_rows(lapply(seq_len(nrow(ftmic_spec)), function(i) {
  r <- ftmic_spec[i, ]
  grid <- seq(0, tau, by = 0.002)
  dplyr::bind_rows(
    tibble::tibble(id = 800L + i, time = 0, evid = 1L, amt = 15 * 25,
                   dur = r$dur, cmt = "central"),
    tibble::tibble(id = 800L + i, time = grid, evid = 0L, amt = NA_real_,
                   dur = 0, cmt = "central")
  ) |>
    dplyr::mutate(label = r$renal, arm = r$inf_label, WT = 25, AGE = 8,
                  RENALIMP_MILD = r$mild, RENALIMP_MOD = r$mod,
                  RENALIMP_SEV = r$sev)
}))

ftmic_sim <- rxode2::rxSolve(mod, events = ftmic_events,
                             keep = c("label", "arm")) |>
  as.data.frame() |>
  dplyr::mutate(label = as.character(label), arm = as.character(arm))

mic_panel <- c(1, 2, 4)

ftmic <- lapply(mic_panel, function(mic) {
  ftmic_sim |>
    dplyr::group_by(label, arm) |>
    dplyr::summarise(
      MIC = mic,
      # Fraction of the 6 h interval with free concentration above the MIC,
      # on a uniform 0.002 h grid.
      pct_fT_over_MIC = 100 * mean(Cfree > mic),
      .groups = "drop"
    )
}) |>
  dplyr::bind_rows()

ftmic |>
  tidyr::pivot_wider(names_from = arm, values_from = pct_fT_over_MIC) |>
  dplyr::arrange(MIC, match(label, c("Normal", "Mild RI", "Moderate RI", "Severe RI"))) |>
  dplyr::rename("Renal function" = label, "MIC (mg/L)" = MIC,
                "%fT>MIC, 0.5 h infusion" = `0.5 h`,
                "%fT>MIC, 3 h infusion" = `3 h`) |>
  knitr::kable(digits = 1,
               caption = paste(
                 "Deterministic %fT>MIC for the typical 8-year-old child,",
                 "15 mg/kg q6h. Target is 70%fT>MIC."))

gain <- ftmic |>
  tidyr::pivot_wider(names_from = arm, values_from = pct_fT_over_MIC) |>
  dplyr::mutate(gain = `3 h` - `0.5 h`)

# The paper's qualitative PD claim: a 3 h infusion raises %fT>MIC in every
# renal stratum and at every MIC in the panel. Deterministic, so exact.
stopifnot(all(gain$gain > 0))
# And the gain is large, not marginal -- at least 15 percentage points
# everywhere, which is what makes it a clinically meaningful recommendation.
stopifnot(min(gain$gain) > 15)
# Severity ordering: worse renal function always gives higher %fT>MIC.
sev_order <- ftmic |>
  dplyr::filter(MIC == 4) |>
  dplyr::arrange(match(label, c("Normal", "Mild RI", "Moderate RI", "Severe RI")))
stopifnot(all(diff(sev_order$pct_fT_over_MIC[sev_order$arm == "3 h"]) > 0))
```

# Assumptions and deviations

* **The whole-body PBPK is not reproduced.** GastroPlus's organ volumes, blood
  flows, tissue-to-plasma partition coefficients and ACAT structure are vendor
  intellectual property and are not printed in the paper. This file encodes the
  reduced disposition model of Supplementary Table S3, which the identity check
  above shows is mono-exponential, and which is what the paper itself carries
  into its dose-adjustment analysis.

* **Supplementary Table S3's volume unit is wrong.** The column is headed
  `Vss(L/kg)` and the Results text repeats "3.259 L/kg in children ... 11.663
  L/kg in adults". The values are absolute litres. Three independent proofs are
  in the identity section above: the published half-lives are recovered only
  with `Vss` in litres; `Vss / body weight` is then a physiological 0.155-0.173
  L/kg; and the litre reading reproduces the paper's own predicted
  concentrations. The model carries the litre reading.

* **The volume is a derived summary, not a transcription.** `lvcPerKg` = 0.1636
  L/kg is the mean of `Vss / WT` over the seven Supplementary Table S3 rows with
  a published weight. It is confirmed independently by solving Table 3's four
  school-age (AUC, Cmax) pairs, which give 0.1629, 0.1635, 0.1659 and 0.1652
  L/kg. Because renal impairment moves this quantity by less than the spread
  between healthy adult cohorts, the model carries no renal effect on volume;
  the 11.663-to-14.081 L range across the adult renal strata is body weight, not
  impairment.

* **Adult clearance is not weight-scaled, deliberately.** Supplementary Table S3
  gives 10.726-11.262 L/h across adult cohorts spanning 58.75-86.6 kg -- flat to
  within 5% over a 47% weight range -- because the GastroPlus renal term is
  driven by glomerular filtration rather than by size. Imposing an allometric
  exponent would have been an invention. Paediatric clearance *is* weight-scaled
  because paediatric dosing and reporting are per kg.

* **Instantaneous-bolus Cmax is not reproduced, and should not be.** The two
  Norrby 1983 rows are simulated in GastroPlus as instantaneous intravenous
  injections, so the whole-body model puts the entire dose in the plasma
  sub-volume before distribution and predicts 165.99 and 331.99 mg/L. A
  one-compartment reduction whose volume is `Vss` cannot produce that, and gives
  roughly a quarter of it. This is the expected `Vss` is not `Vc` failure. It is
  not a defect of the reduction for this paper's purposes: Table 2 reports **no
  observed Cmax** for either bolus row, no clinical regimen in the paper is an
  instantaneous injection, and both bolus AUCs are reproduced to within 2%.
  The Gibson 0.15 h infusions are the same effect in milder form (fold error
  0.72-0.81) and likewise carry no observed Cmax. The effect disappears as soon
  as the infusion is long enough to matter: the 0.5 h, 1 h, 2 h and 3 h
  regimens -- which is every regimen the paper recommends -- are reproduced to
  within 10%, and the four school-age rows to within 1%.

* **Interindividual variability and residual error are not published.** Feng
  2026 ran a 200-subject population simulator and shows 90% intervals in Figures
  2-4, but reports no variance magnitude for any parameter and describes no
  residual error model. No etas are declared and both residual error terms are
  `fixed(0)` rather than invented. The direct consequence is that the paper's
  probability-of-target-attainment results (Figure 6) **cannot be reproduced**,
  because PTA is defined over that unpublished distribution. The deterministic
  %fT>MIC computed above is the closest reproducible quantity, and it sits below
  the paper's PTA percentages -- as it must, since a PTA is the fraction of a
  population above a threshold, not a typical value.

* **Paediatric renal-impairment unbound fraction is not published.** Supplementary
  Table S3 carries `Fup` for adults in every renal stratum (0.800, 0.806, 0.827,
  0.834) but only for children with *normal* renal function (0.81217 Claesson,
  0.81182 Bradley). The model therefore applies the paediatric normal-renal value
  to children in every renal stratum. The adult trend suggests the true paediatric
  RI values would be at most about 0.834, a 2.7% effect on the free concentration.

* **One printed geometric mean ratio disagrees with its own AUC pair.** Table 4
  gives 1.19 as the moderate-RI-to-healthy ratio for preschool children, but the
  AUC values in the same column pair are 38.928 and 31.358, whose ratio is
  1.2414 -- 4.3% apart. The other eight ratios agree with their AUC pairs to
  within 0.9%. The model derives the clearance multipliers from the tabulated
  AUC values rather than the tabulated ratios. The discrepancy does not change
  the paper's recommendation: 15 / 1.19 and 15 / 1.2414 both round to 12 mg/kg.
  Separately, Table 4 prints 1.25 for school-age moderate RI while the Abstract
  and Discussion print 1.26 for the same quantity; 53.765 / 42.818 = 1.2557, so
  the Abstract is right and Table 4 is truncated.

* **One predicted adult AUC exceeds what the paper's own clearance allows.**
  Table 2 gives Jaruratanasirikul 2005 two 0.5 g regimens, a 0.5 h and a 2.0 h
  infusion, and Supplementary Table S3 gives both the same clearance of 10.726
  L/h. A 500 mg dose at that clearance cannot produce an AUC above
  500 / 10.726 = 46.62 mg*h/L, and the AUC0-t the table footnote defines is a
  *truncated* AUC, so it must be below that. The 2.0 h row obeys this (46.20).
  The 0.5 h row does not: it is printed as 50.91, which is 9.2% above the
  theoretical maximum. This is the single regimen where the reduction's AUC
  ratio falls outside 5% (0.879), and the discrepancy is in the source, not in
  the reduction -- the reduction returns 46.62, which is the value Table 2's own
  inputs imply.

* **The Bradley 2023 simulation does not use the Bradley 2023 cohort.**
  Supplementary Table S2 gives that cohort a mean weight of 15.4 kg (range 12-19)
  at ages 2-6, so a 15 mg/kg dose is 231 mg. Supplementary Table S3 gives the
  Bradley simulation a clearance of 8.489 L/h, which would cap AUC0-inf at
  231 / 8.489 = 27.2 mg*h/L -- yet Table 3 reports a predicted AUC0-t of 47.40,
  which is impossible for a truncated AUC. Solving Table 3's Bradley AUC and
  Cmax jointly implies a subject of about 26 kg, i.e. a school-age child rather
  than the 2-6 year olds Bradley enrolled. This vignette does not simulate the
  Bradley regimen against the reported 15.4 kg for that reason; the paediatric
  arm uses the Table 4 age bands, which are internally consistent.

* **Cross-row identity in Supplementary Table S3.** The Nilsson 1991 row (Swedish
  men, 74 kg) and the Wang 2021 q6h row (Chinese participants, 60.4 kg, 41.7%
  male) share an identical `Vss` of 11.663 L, `GFR` of 1.9632 mL/s, half-life of
  0.723 h and renal blood flow of 21.35 mL/s. Several study rows were evidently
  simulated with the same default virtual adult rather than with each study's own
  demographics, which is why the model carries a single adult clearance rather
  than a per-study one.

* **Clearance is not the paper's stated `Fup * GFR`.** Methods and Discussion both
  state that in vivo renal clearance was computed as `Fup * GFR` with no non-renal
  pathway. Supplementary Table S3's own columns do not satisfy that relation: for
  the Nilsson row, `Fup * GFR` is 0.8 * 1.9632 mL/s = 0.8 * 7.068 L/h = 5.65 L/h
  against a tabulated clearance of 11.126 L/h, and the ratio of tabulated
  clearance to `Fup * GFR` varies from 1.95-2.34 across the healthy adult and
  paediatric rows to 4.8 in moderate RI and 8.1 in severe RI. The model carries the
  tabulated clearances, which are the numbers that reproduce the paper's published
  concentrations, rather than the stated formula.

* **Ages below 3 years produce a clearance of zero.** The WHO bands the model
  uses start at 3 years, matching the paper's own Limitations statement that the
  work applies only to children over 3 with normal growth and development.
  An out-of-domain age yields nonsense rather than an error, so callers must
  screen for it.
