---
title: "Ipilimumab tumor growth dynamics (Feng 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ipilimumab tumor growth dynamics (Feng 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Feng_2019_ipilimumab"))$reference`
* Description: `r rxode2::rxode(readModelDb("Feng_2019_ipilimumab"))$description`
* Article: <https://doi.org/10.1002/psp4.12454> (open access, CC BY-NC)
* Supplement: Figures S1-S6 and Tables S1-S5, distributed with the article on the
  CPT:PSP website and mirrored in PubMed Central under
  [PMC6875707](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6875707/)

Feng 2019 develops a **mixture tumor growth dynamics (TGD) model** for tumor
burden in 688 patients with advanced melanoma from the phase III study CA184-169
(NCT01515189), and then links a derived measure of tumor response to overall
survival through a Cox proportional-hazards model.

This package carries the **TGD model** (the paper's TGD-Model 11, the final
covariate mixture model). The overall-survival model is *not* packaged, for the
reason set out under [Assumptions and deviations](#assumptions-and-deviations).

# Population

```{r population}
pop <- readModelDb("Feng_2019_ipilimumab")()$population
tibble::tibble(
  Field = names(pop),
  Value = vapply(pop, function(x) paste(as.character(x), collapse = "; "), character(1))
) |>
  knitr::kable()
```

Patients received ipilimumab 3 mg/kg (n = 343) or 10 mg/kg (n = 345) as a
90-minute intravenous infusion every 3 weeks for four doses. Tumor burden (TB)
is the sum of the longest diameters of target lesions, in cm, assessed under
immune-related response criteria at weeks 12, 16 and 24.

# Source trace

Every packaged value, traced to its location in Feng 2019.

```{r source-trace, echo = FALSE}
tribble(
  ~Quantity,                            ~Value,      ~`Source location`,
  "TB(t) = TB0*exp(-TS*t) + TG*t + TBss", "equation",  "Methods, 'TGD model' -- the three per-subpopulation structural equations",
  "P_i = P_TV * exp(eta_i) (log-normal IIV)", "equation", "Methods, 'TGD model', interindividual-variability equation",
  "TB_obs = TB*(1 + propSd*eps) + addSd*eps", "equation", "Methods, 'TGD model', residual-error equation (single eps)",
  "TB0 fast",                           "10.2 cm",   "Table 2, Fixed effects / Fast / TB0",
  "TG fast",                            "0.328 cm/wk", "Table 2, Fixed effects / Fast / TG",
  "TS fast",                            "0.00362 /wk", "Table 2, Fixed effects / Fast / TS",
  "TB0 no-growth",                      "2.53 cm",   "Table 2, Fixed effects / No growth / TB0",
  "TS no-growth",                       "0.0458 /wk", "Table 2, Fixed effects / No growth / TS",
  "TBss no-growth",                     "1.71 cm",   "Table 2, Fixed effects / No growth / TBss",
  "TB0 intermediate",                   "5.83 cm",   "Table 2, Fixed effects / Intermediate / TB0 P3",
  "TG intermediate",                    "0.0236 cm/wk", "Table 2, Fixed effects / Intermediate / TG P3",
  "TS intermediate",                    "0.00299 /wk", "Table 2, Fixed effects / Intermediate / TS P3",
  "TP1 / TP2 (mixture weights)",        "1.20 / 0.878", "Table 2 and footnote e",
  "LDH effect on TB0",                  "0.868",     "Table 2, Fixed effects / LDH effect on TB0",
  "LDH effect on TG",                   "0.771",     "Table 2, Fixed effects / LDH effect on TG",
  "Cavg1 effect on TG",                 "-0.00342",  "Table 2, Fixed effects / Exposure (Cavg1) effect on TG",
  "Covariate functional form",          "(1 + log(BLDHU)*coef), (1 + Cavg1*coef)", "Table 2 footnote f",
  "omega^2 TB0 (shared)",               "0.535",     "Table 2, Random effects / omega^2 TB0",
  "omega^2 TG / TS fast, covariance",   "0.360 / 4.07, -1.06", "Table 2, Random effects / Fast",
  "omega^2 TS / TBss no-growth",        "0.385 / 1.38", "Table 2, Random effects / No growth",
  "omega^2 TG / TS interm., covariance", "0.203 / 3.21, 0.129", "Table 2, Random effects / Intermediate",
  "Additive residual error",            "0.125 cm",  "Table 2, Residual error / Additive error",
  "Proportional residual error",        "0.167",     "Table 2, Residual error / Proportional error",
  "Final model identification",         "TGD-Model 11", "Table S1 (BIC 6032.321, the reference model)",
  "Subpopulation fractions by arm",     "see below", "Table S2",
  "Demographics",                       "see above", "Table 1"
) |>
  knitr::kable()
```

# Model structure

The published model is **algebraic**, not an ODE system. Tumor burden for
patient *i* at time *t* (weeks since first dose) is

$$TB_i(t) = TB0_i \cdot e^{-TS_i t} + TG_i \cdot t + TBss_i$$

where the three latent subpopulations differ in which terms are active:

| Subpopulation | `MIX_FAST_GROW` | `MIX_NO_GROW` | Active terms |
|---|---|---|---|
| Fast tumor growth | 1 | 0 | TB0, TS, TG (TBss structurally 0) |
| No growth | 0 | 1 | TB0, TS, TBss (TG structurally 0) |
| Intermediate TG and TS | 0 | 0 | TB0, TS, TG (TBss structurally 0) |

Feng 2019 added the `TBss` plateau term to the pre-immunotherapy Wang model
specifically to capture *"TB time profiles that asymptotically approach a
steady-state value -- an immunotherapy-specific response pattern observed in
some patients"* (Methods). Table S1 confirms this mattered: dropping `TBss`
(TGD-Model 15, "original Wang model without TBss") raised BIC by 217.5 units.

## Mixture proportions

Table 2 footnote e defines the class probabilities from `TP1` and `TP2`.

```{r mixture-probs}
mod <- rxode2::rxode(readModelDb("Feng_2019_ipilimumab"))
tp1 <- mod$theta[["tp1"]]
tp2 <- mod$theta[["tp2"]]

probs <- c(
  `Fast TG`               = tp1 / (1 + tp1 + tp2),
  `No growth`             = 1   / (1 + tp1 + tp2),
  `Intermediate TG & TS`  = tp2 / (1 + tp1 + tp2)
)
round(100 * probs, 1)

# The three probabilities must sum to exactly 1 (Table 2 footnote e:
# "The sum of overall probabilities was 1").
stopifnot(abs(sum(probs) - 1) < 1e-12)
```

Feng 2019 **Table S2** reports the *empirical* per-arm subpopulation membership:

```{r tableS2}
tableS2 <- tribble(
  ~Arm,        ~`No growth`, ~`Intermediate TG & TS`, ~`Fast TG`, ~n,
  "3 mg/kg",   24.2,         28.9,                    46.9,       343L,
  "10 mg/kg",  28.7,         30.4,                    40.9,       345L
)
knitr::kable(tableS2)

# Pooled across arms, weighted by arm size.
pooled <- with(tableS2, c(
  `Fast TG`              = sum(`Fast TG` * n) / sum(n),
  `No growth`            = sum(`No growth` * n) / sum(n),
  `Intermediate TG & TS` = sum(`Intermediate TG & TS` * n) / sum(n)
))
comp <- tibble::tibble(
  Subpopulation = names(probs),
  `Model probability (%)` = round(100 * probs[names(probs)], 1),
  `Table S2 pooled (%)`   = round(pooled[names(probs)], 1)
)
knitr::kable(comp)

# The two columns measure different things (see Errata): the model column is
# the prior class probability from TP1/TP2, the Table S2 column is the pooled
# per-subject maximum-posterior assignment. They are NOT expected to be
# identical, so this is a sanity bound, not an identity check.
maxGap <- max(abs(100 * probs[names(probs)] - pooled[names(probs)]))
round(maxGap, 2)

stopifnot(
  # Fast TG is the largest class on both readings.
  which.max(probs) == which.max(pooled[names(probs)]),
  # No class disagrees by more than 8 percentage points.
  maxGap < 8
)
```

The largest disagreement between the two readings is
`r round(maxGap, 1)` percentage points.

The paper's headline dose finding is visible directly in Table S2: the 10 mg/kg
arm has a **higher** fraction in the no-growth class (28.7% vs 24.2%) and a
**lower** fraction in the fast-growth class (40.9% vs 46.9%).

```{r tableS2-direction}
stopifnot(
  tableS2$`No growth`[tableS2$Arm == "10 mg/kg"] > tableS2$`No growth`[tableS2$Arm == "3 mg/kg"],
  tableS2$`Fast TG`[tableS2$Arm == "10 mg/kg"]   < tableS2$`Fast TG`[tableS2$Arm == "3 mg/kg"]
)
```

# Typical-value profiles by subpopulation

A one-subject-per-subpopulation ladder with random effects zeroed reproduces the
three qualitative patterns the paper describes.

```{r typical-ladder}
modTypical <- rxode2::zeroRe(mod)

subpops <- tibble::tibble(
  id            = 1:3,
  Subpopulation = c("Fast TG", "No growth", "Intermediate TG & TS"),
  MIX_FAST_GROW = c(1, 0, 0),
  MIX_NO_GROW   = c(0, 1, 0)
)

tgrid <- seq(0, 52, by = 1)
evTypical <- subpops |>
  tidyr::crossing(time = tgrid) |>
  dplyr::mutate(LDHR = 1, CAV = 0) |>
  dplyr::arrange(id, time) |>
  as.data.frame()

simTypical <- as.data.frame(rxode2::rxSolve(modTypical, evTypical)) |>
  dplyr::left_join(subpops[, c("id", "Subpopulation")], by = "id")

# Structural identity: the solved profile must equal the published closed form
# evaluated with the same parameters. Both sides use identical drawn values, so
# the only difference is floating-point noise and a tight bound is correct.
chkClosed <- simTypical |>
  dplyr::mutate(closed_form = tb0 * exp(-ts * time) + tg * time + tbss)
stopifnot(max(abs(chkClosed$tumor_size - chkClosed$closed_form)) < 1e-10)

# Published typical values are recovered exactly at their own anchors.
anchors <- simTypical |>
  dplyr::filter(time == 0) |>
  dplyr::select(Subpopulation, tb0, ts, tg, tbss)
knitr::kable(anchors, digits = 5)
stopifnot(
  isTRUE(all.equal(anchors$tb0,  c(10.2, 2.53, 5.83),         tolerance = 1e-6)),
  isTRUE(all.equal(anchors$ts,   c(0.00362, 0.0458, 0.00299), tolerance = 1e-6)),
  isTRUE(all.equal(anchors$tg,   c(0.328, 0, 0.0236),         tolerance = 1e-6)),
  isTRUE(all.equal(anchors$tbss, c(0, 1.71, 0),               tolerance = 1e-6))
)
```

```{r typical-plot, fig.cap = "Typical-value tumor-burden profiles for the three mixture subpopulations. Compare with the three response patterns described in Feng 2019 Results and shown per-individual in Figure 2b."}
ggplot(simTypical, aes(time, tumor_size, colour = Subpopulation)) +
  geom_line(linewidth = 1) +
  labs(x = "Time since first dose (weeks)", y = "Tumor burden (cm)",
       colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## The no-growth class reaches a plateau; the others do not

```{r plateau-check}
late <- simTypical |> dplyr::filter(time == 52)
nogrow <- late[late$Subpopulation == "No growth", ]

# TB(t) -> TBss as t -> infinity in the no-growth class. At t = 52 weeks the
# decaying component has fallen to exp(-0.0458*52) = 9.2% of TB0.
stopifnot(
  nogrow$tumor_size > nogrow$tbss,
  (nogrow$tumor_size - nogrow$tbss) / nogrow$tbss < 0.15
)

# Tumor burden is monotone DECREASING throughout for the no-growth class
# (dTB/dt = -TB0*TS*exp(-TS*t) < 0 for all t), and monotone INCREASING for the
# fast class.
mono <- simTypical |>
  dplyr::group_by(Subpopulation) |>
  dplyr::summarise(all_down = all(diff(tumor_size) < 0),
                   all_up   = all(diff(tumor_size) > 0), .groups = "drop")
knitr::kable(mono)
stopifnot(
  mono$all_down[mono$Subpopulation == "No growth"],
  mono$all_up[mono$Subpopulation == "Fast TG"]
)
```

Note that at the *typical value* the intermediate class is monotonically
increasing too: its initial slope is
`-5.83 * 0.00299 + 0.0236 = +0.0062` cm/week. The "initial shrinkage followed by
tumor progression" pattern the paper attributes to this class is driven by its
very large shrinkage-rate variability (omega^2 TS = 3.21, SD 1.79 on the log
scale), not by the typical profile -- quantified in the cohort section below.

# Deriving the exposure covariate Cavg1

The TGD model takes `CAV` (the paper's Cavg1, time-averaged concentration after
the first dose) as an input. Feng 2019 obtained it from a separate ipilimumab
population PK analysis; here it is derived from the companion model
`Feng_2014_ipilimumab`, from the same Bristol-Myers Squibb group and the same
drug, using **PKNCA** for the interval AUC.

```{r cavg1}
pkMod <- rxode2::zeroRe(rxode2::rxode(readModelDb("Feng_2014_ipilimumab")))
tauDays <- 21   # q3w, and the Feng 2014 model runs on a day time base

wtRef <- 79.9   # Feng 2019 Table 1 overall mean body weight
ldhRefPk <- 206 # Feng 2014 reference LDH (U/L)

pkConc <- lapply(c(3, 10), function(mgkg) {
  ev <- rxode2::et(amt = mgkg * wtRef, dur = 1.5 / 24, cmt = "central") |>
    rxode2::et(seq(0, tauDays, length.out = 505))
  as.data.frame(rxode2::rxSolve(pkMod, ev, params = c(WT = wtRef, LDH = ldhRefPk))) |>
    dplyr::transmute(id = paste0(mgkg, " mg/kg"), time, conc = Cc, mgkg = mgkg)
}) |>
  dplyr::bind_rows() |>
  dplyr::filter(!is.na(conc))

pkDose <- pkConc |>
  dplyr::distinct(id, mgkg) |>
  dplyr::mutate(time = 0, amount = mgkg * wtRef)

concObj <- PKNCA::PKNCAconc(pkConc, conc ~ time | id)
doseObj <- PKNCA::PKNCAdose(pkDose, amount ~ time | id)
intervals <- data.frame(start = 0, end = tauDays, auclast = TRUE, cmax = TRUE, tmax = TRUE)
ncaRes <- PKNCA::pk.nca(PKNCA::PKNCAdata(concObj, doseObj, intervals = intervals))

cavg1 <- as.data.frame(ncaRes) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::transmute(id, cavg1 = PPORRES / tauDays)
cavg1 |>
  dplyr::mutate(
    `TG multiplier` = 1 + cavg1 * mod$theta[["e_cav_tg"]],
    cavg1 = round(cavg1, 1),
    `TG multiplier` = round(`TG multiplier`, 3)
  ) |>
  dplyr::rename("Dose arm" = id, "Cavg1 (ug/mL)" = cavg1) |>
  knitr::kable()
```

Two things follow, and both are load-bearing.

**First, the units of Cavg1.** Feng 2019 never states them. The Table 2
coefficient enters as the linear multiplier `1 + Cavg1 * (-0.00342)`, which is
positive only for Cavg1 < 292. The derived values above are ~21 and ~69, giving
multipliers safely inside `(0, 1)`. Read as ng/mL the same exposures would give
a multiplier near -234 -- a negative tumor-growth rate of absurd magnitude; read
as mg/mL the covariate effect would vanish to four significant figures,
contradicting its retention in the final model. **ug/mL is the only
dimensionally feasible reading**, and it is also the concentration unit used by
both registry ipilimumab popPK models from the same group.

**Second, the direction of the dose effect.** The paper's central exposure
finding is a *lower* tumor growth rate at 10 mg/kg.

```{r cavg1-direction}
mult <- setNames(1 + cavg1$cavg1 * mod$theta[["e_cav_tg"]], cavg1$id)
stopifnot(
  # Both multipliers are physically valid (a positive growth rate).
  all(mult > 0),
  # Higher dose -> lower growth rate (Feng 2019 abstract and Results).
  mult[["10 mg/kg"]] < mult[["3 mg/kg"]]
)
```

# Virtual cohort

200 subjects per dose arm (the per-arm cap), with subpopulation drawn from the
model's own class probabilities.

```{r cohort}
rxode2::rxSetSeed(20190001)
set.seed(20190001)

nPerArm <- 200L
arms <- c("3 mg/kg", "10 mg/kg")

cohort <- tidyr::expand_grid(Arm = arms, subj = seq_len(nPerArm)) |>
  dplyr::mutate(
    id  = dplyr::row_number(),
    cls = sample(names(probs), dplyr::n(), replace = TRUE, prob = probs),
    MIX_FAST_GROW = as.integer(cls == "Fast TG"),
    MIX_NO_GROW   = as.integer(cls == "No growth"),
    # LDHR: log-normal with median exactly 1.0 (Feng 2019 Table 1 median),
    # truncated to the reported 0.4-40.5 range. The paper reports only the
    # median and range, so the shape is an assumption -- see Errata.
    LDHR = pmin(pmax(stats::rlnorm(dplyr::n(), meanlog = 0, sdlog = 0.6), 0.4), 40.5)
  ) |>
  dplyr::left_join(
    tibble::tibble(Arm = cavg1$id, CAV = cavg1$cavg1),
    by = "Arm"
  ) |>
  dplyr::select(id, Arm, Subpopulation = cls, MIX_FAST_GROW, MIX_NO_GROW, LDHR, CAV)

# Sanity: LDHR median is the published 1.0, and every subject is in exactly one class.
stopifnot(
  abs(stats::median(cohort$LDHR) - 1) < 0.12,
  all(cohort$MIX_FAST_GROW + cohort$MIX_NO_GROW <= 1L)
)

events <- cohort |>
  tidyr::crossing(time = tgrid) |>
  dplyr::arrange(id, time) |>
  as.data.frame()

sim <- as.data.frame(rxode2::rxSolve(mod, events)) |>
  dplyr::left_join(cohort[, c("id", "Arm", "Subpopulation")], by = "id")
```

## Baseline tumor burden reproduces Table 1

This is the strongest available quantitative check on the packaged values: the
cohort mean baseline tumor burden depends jointly on all three `TB0` anchors,
the `TBss` anchor, the mixture probabilities, and the IIV variances. Feng 2019
Table 1 reports a mean baseline TB of **9.6 cm (SD 8.7)** over the whole
analysis population.

Note that in the no-growth class the model's baseline is `TB0 + TBss` (at
*t* = 0 the exponential term equals `TB0` and the plateau term is already
present), not `TB0` alone.

```{r baseline-check}
baseline <- sim |> dplyr::filter(time == 0)

# Baseline is TB0 + TBss by construction; confirm the model agrees.
stopifnot(max(abs(baseline$tumor_size - (baseline$tb0 + baseline$tbss))) < 1e-10)

baselineSummary <- tibble::tibble(
  Source = c("Feng 2019 Table 1 (observed)", "Simulated cohort"),
  `Mean baseline TB (cm)` = c(9.6, round(mean(baseline$tumor_size), 2)),
  `SD (cm)` = c(8.7, round(stats::sd(baseline$tumor_size), 2))
)
knitr::kable(baselineSummary)

# Assert on the CENTRE of the distribution, which is what the packaged anchors
# and mixture weights determine. The mean of the analytic mixture is
#   sum_k p_k * E[TB0_k * exp(eta_TB0) + TBss_k * exp(eta_TBss)]
# with E[exp(eta)] = exp(omega^2 / 2); evaluated below independently of the
# simulation as a second, sampling-free check.
oTb0  <- 0.535
oTbss <- 1.38
analyticMean <-
  probs[["Fast TG"]]              * 10.2 * exp(oTb0 / 2) +
  probs[["Intermediate TG & TS"]] * 5.83 * exp(oTb0 / 2) +
  probs[["No growth"]]            * (2.53 * exp(oTb0 / 2) + 1.71 * exp(oTbss / 2))
round(analyticMean, 2)

stopifnot(
  # Analytic mixture mean lands within 10% of the published 9.6 cm.
  abs(analyticMean - 9.6) / 9.6 < 0.10,
  # And the simulated cohort agrees with the analytic mean.
  abs(mean(baseline$tumor_size) - analyticMean) / analyticMean < 0.20
)
```

The analytic mixture mean is `r round(analyticMean, 2)` cm against a published
`9.6` cm -- agreement to within
`r round(100 * abs(analyticMean - 9.6) / 9.6, 1)`%. Because this single number
is a weighted combination of every baseline anchor, the `TBss` plateau, the
mixture weights and two IIV variances, a transcription error in any one of them
would move it noticeably.

## Initial shrinkage in the intermediate class is IIV-driven

```{r initial-slope}
slope0 <- baseline |>
  dplyr::mutate(dTB_dt0 = -tb0 * ts + tg) |>
  dplyr::group_by(Subpopulation) |>
  dplyr::summarise(`% with initial shrinkage` = round(100 * mean(dTB_dt0 < 0), 1),
                   .groups = "drop")
knitr::kable(slope0)

# Every no-growth subject shrinks initially (TG is structurally zero, so
# dTB/dt = -TB0*TS < 0 always). A substantial minority of intermediate-class
# subjects do too, which is the "initial shrinkage followed by tumor
# progression" pattern Feng 2019 Results ascribes to that class -- even though
# the TYPICAL intermediate profile is monotonically increasing.
stopifnot(
  slope0$`% with initial shrinkage`[slope0$Subpopulation == "No growth"] == 100,
  slope0$`% with initial shrinkage`[slope0$Subpopulation == "Intermediate TG & TS"] > 20
)
```

## Replicating Figure 2: change in tumor burden from baseline

```{r figure2, fig.height = 5.5, fig.cap = "Simulated percentage change in tumor burden from baseline, by dose arm and mixture subpopulation. Replicates the structure of Feng 2019 Figure 2b (individual model-predicted time course, stratified by dose and subpopulation)."}
pctChange <- sim |>
  dplyr::group_by(id) |>
  dplyr::mutate(pct = 100 * (tumor_size / tumor_size[time == 0] - 1)) |>
  dplyr::ungroup()

ggplot(pctChange, aes(time, pct, group = id)) +
  geom_line(alpha = 0.18, linewidth = 0.3) +
  stat_summary(aes(group = 1), fun = median, geom = "line",
               colour = "firebrick", linewidth = 1) +
  facet_grid(Arm ~ Subpopulation) +
  coord_cartesian(ylim = c(-100, 300)) +
  labs(x = "Time since first dose (weeks)",
       y = "Change in tumor burden from baseline (%)") +
  theme_bw()
```

## Replicating Figure 1: parameter distributions by subpopulation

Feng 2019 Figure 1 shows the distributions of the TGD parameters by
subpopulation, and the Results state two specific orderings: *"TB0 was higher in
patients with fast TG, and TS was higher in patients with no-growth compared
with other subpopulations"*, and *"Patients in the fast TG subpopulation had
higher TG relative to patients in the intermediate TG and TS group"*.

```{r figure1, fig.height = 3.6, fig.cap = "Distribution of individual TGD parameters by subpopulation. Replicates Feng 2019 Figure 1."}
paramLong <- baseline |>
  dplyr::select(Subpopulation, TB0 = tb0, TS = ts, TG = tg) |>
  tidyr::pivot_longer(c(TB0, TS, TG), names_to = "Parameter", values_to = "Value") |>
  # TG is structurally zero in the no-growth class, which has no place on a log
  # axis; drop those rows rather than let them become -Inf.
  dplyr::filter(Value > 0)

ggplot(paramLong, aes(Subpopulation, Value, fill = Subpopulation)) +
  geom_boxplot(outlier.size = 0.4) +
  facet_wrap(~Parameter, scales = "free_y") +
  scale_y_log10() +
  labs(x = NULL, y = "Individual parameter value (log scale)") +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = "bottom")
```

```{r figure1-orderings}
med <- baseline |>
  dplyr::group_by(Subpopulation) |>
  dplyr::summarise(TB0 = stats::median(tb0), TS = stats::median(ts),
                   TG = stats::median(tg), .groups = "drop")
knitr::kable(med, digits = 5)

getMed <- function(col, cls) med[[col]][med$Subpopulation == cls]

stopifnot(
  # "TB0 was higher in patients with fast TG"
  getMed("TB0", "Fast TG") > getMed("TB0", "No growth"),
  getMed("TB0", "Fast TG") > getMed("TB0", "Intermediate TG & TS"),
  # "TS was higher in patients with no-growth compared with other subpopulations"
  getMed("TS", "No growth") > getMed("TS", "Fast TG"),
  getMed("TS", "No growth") > getMed("TS", "Intermediate TG & TS"),
  # "Patients in the fast TG subpopulation had higher TG relative to patients
  # in the intermediate TG and TS group"
  getMed("TG", "Fast TG") > getMed("TG", "Intermediate TG & TS")
)
```

# Tumor-response measures and the link to survival

Feng 2019 derives four measures of tumor response from the TGD model and selects
**PRW8** (progression rate at week 8, the first derivative of tumor burden with
respect to time at week 8) as the best single predictor of overall survival.
`CTB8` is the relative change in tumor burden at week 8, as a percentage of
baseline.

```{r prw8}
week8 <- sim |> dplyr::filter(time == 8)

resp <- week8 |>
  dplyr::left_join(dplyr::select(baseline, id, tb_base = tumor_size), by = "id") |>
  dplyr::mutate(
    # PRW8 = dTB/dt at t = 8 = -TB0*TS*exp(-TS*8) + TG
    PRW8 = -tb0 * ts * exp(-ts * 8) + tg,
    CTB8 = 100 * tumor_size / tb_base
  )

# Identity check: the analytic derivative must match a central finite difference
# of the solved profile. Both sides use the same drawn parameters, so this is
# pure numerical error and a tight bound is correct.
h <- 1e-4
fd <- sim |>
  dplyr::filter(time == 8) |>
  dplyr::transmute(id,
                   fd = ((tb0 * exp(-ts * (8 + h)) + tg * (8 + h) + tbss) -
                         (tb0 * exp(-ts * (8 - h)) + tg * (8 - h) + tbss)) / (2 * h))
chkD <- dplyr::left_join(resp[, c("id", "PRW8")], fd, by = "id")
stopifnot(max(abs(chkD$PRW8 - chkD$fd)) < 1e-6)

respSummary <- resp |>
  dplyr::group_by(Subpopulation) |>
  dplyr::summarise(`Median PRW8 (cm/week)` = round(stats::median(PRW8), 4),
                   `Median CTB8 (% of baseline)` = round(stats::median(CTB8), 1),
                   .groups = "drop")
knitr::kable(respSummary)
```

Feng 2019 Results state that *"the risk of death increased with an increase in
PRW8.MIXC"*, and supplement **Table S5** reports the hazard ratios between
subpopulations with fast TG as reference: intermediate 0.53 (95% PI 0.42-0.71)
and no-growth 0.33 (0.25-0.44). The TGD model and the survival model must
therefore agree on the ordering: the subpopulation with the highest PRW8 must be
the one with the highest hazard.

```{r prw8-ordering}
getResp <- function(col, cls) respSummary[[col]][respSummary$Subpopulation == cls]

tableS5 <- tibble::tibble(
  Subpopulation = c("Fast TG", "Intermediate TG & TS", "No growth"),
  `Table S5 HR vs fast TG` = c(1.00, 0.53, 0.33),
  `Median PRW8 (cm/week)` = c(getResp("Median PRW8 (cm/week)", "Fast TG"),
                              getResp("Median PRW8 (cm/week)", "Intermediate TG & TS"),
                              getResp("Median PRW8 (cm/week)", "No growth"))
)
knitr::kable(tableS5)

# PRW8 must decrease in exactly the order the hazard ratios decrease.
stopifnot(
  !is.unsorted(rev(tableS5$`Median PRW8 (cm/week)`)),
  identical(order(tableS5$`Median PRW8 (cm/week)`, decreasing = TRUE),
            order(tableS5$`Table S5 HR vs fast TG`, decreasing = TRUE))
)

# And the qualitative reading of CTB8: fast grows past baseline by week 8,
# no-growth has shrunk below it.
stopifnot(
  getResp("Median CTB8 (% of baseline)", "Fast TG")   > 100,
  getResp("Median CTB8 (% of baseline)", "No growth") < 100
)
```

# Covariate effects

Feng 2019 Results: *"Estimated LDH effects on TB0 and TG showed that higher
baseline LDH was associated with a higher TG rate and higher baseline tumor
size."*

```{r ldh-effect}
ldhGrid <- tibble::tibble(id = seq_len(5), LDHR = c(0.5, 1, 2, 5, 10),
                          MIX_FAST_GROW = 1, MIX_NO_GROW = 0, CAV = 0)
ldhEv <- ldhGrid |>
  tidyr::crossing(time = c(0, 8)) |>
  dplyr::arrange(id, time) |>
  as.data.frame()
# rxSolve returns the covariate columns (including LDHR) alongside the model
# quantities, so no join back onto ldhGrid is needed -- and doing one would
# collide on the LDHR name.
ldhSim <- as.data.frame(rxode2::rxSolve(modTypical, ldhEv)) |>
  dplyr::filter(time == 0) |>
  dplyr::arrange(LDHR)

ldhSim |>
  dplyr::transmute(LDHR, `TB0 (cm)` = round(tb0, 2), `TG (cm/week)` = round(tg, 4)) |>
  knitr::kable()

stopifnot(
  # Monotone increasing in LDHR, as the Results sentence requires.
  !is.unsorted(ldhSim$tb0),
  !is.unsorted(ldhSim$tg),
  # LDHR = 1 (patient at the laboratory ULN) is the neutral point: the
  # multiplier (1 + log(1)*coef) is exactly 1, so TB0 equals its anchor.
  abs(ldhSim$tb0[ldhSim$LDHR == 1] - 10.2) < 1e-8
)
```

The exposure effect on TG was retained in the final model but was not
statistically significant -- Feng 2019 Results: *"TG decreased with Cavg1;
however, the 95% CI of parameter estimate included zero"* (Table 2:
-0.00342, 95% CI -0.00690 to 6.69E-05). The Discussion is explicit that *"the
effect of Cavg1 on TGD model parameters was not significant"*, and attributes
the observed dose difference in survival instead to the *class-membership*
difference shown in Table S2. The packaged model reproduces the point estimate;
users should not read it as an established exposure-response relationship.

# Assumptions and deviations {#assumptions-and-deviations}

**The overall-survival model is not packaged.** Feng 2019's OS model is a
*semiparametric* Cox proportional-hazards regression: `lambda(t) = lambda0(t) *
exp(beta' X)`. The nonparametric baseline hazard `lambda0(t)` is neither
reported in the paper nor in the supplement, and by construction has no
closed-form published representation (it is a step function estimated at the
observed event times). Simulating survival times therefore is not possible from
the published material, and no amount of additional source acquisition would
change that. The regression's estimated effects are shown graphically in
Figure 3 and its model-selection statistics in Table S3; the predicted hazard
ratios are in Tables S4 and S5, and Table S5 is used above as a check on the TGD
model's PRW8 ordering. Nothing about the TGD model is lost by this omission --
the TGD model is complete and self-contained.

**Units of Cavg1 (`CAV`) are inferred, not stated.** Feng 2019 gives no unit for
Cavg1 anywhere in the text, tables, or supplement. ug/mL is adopted on the
dimensional-feasibility argument set out in the Cavg1 section above, supported by
both registry ipilimumab popPK models from the same group using ug/mL.

**Erratum: Table 2 footnote f reuses the TG coefficient in the TB0 equation.**
The footnote prints

```
TG_Tv  = TG_REF  * (1 + log(BLDHU) * TG_BLDHU) * (1 + Cavg1 * TG_Cavg1)
TB0_Tv = TB0_REF * (1 + log(BLDHU) * TG_BLDHU)
```

using the symbol `TG_BLDHU` in *both* lines. This is a typographical slip:
Table 2 tabulates two distinct estimates with distinct confidence intervals
("LDH effect on TB0" = 0.868, 95% CI 0.752-0.984; "LDH effect on TG" = 0.771,
95% CI 0.473-1.07), and the Results describe them as two separate findings. The
packaged model uses the TB0 coefficient (0.868) in the TB0 equation.

**Table 2 footnote a does not apply to any row.** Footnote a states that fixed
(non-estimated) parameters are flagged with a superscript "f". Three rows carry
a superscript `f` -- "LDH effect on TB0", "LDH effect on TG" and "Exposure
(Cavg1) effect on TG" -- but that `f` is the *footnote-f* marker pointing at the
covariate-equation footnote, not a fixed flag: all three rows report 95%
confidence intervals, and the Results text calls them estimated. No parameter in
Table 2 is fixed. The only structurally fixed quantities are the zeros the
mixture imposes (TG = 0 in the no-growth class; TBss = 0 in the fast and
intermediate classes), which are encoded structurally rather than as `fixed()`
parameters.

**Model class probabilities vs Table S2 assignments.** The `TP1`/`TP2`
parameters give *prior* class probabilities of 39.0 / 32.5 / 28.5% (fast /
no-growth / intermediate). Table S2 reports 43.9 / 26.5 / 29.7% pooled across
arms. These differ by up to `r round(maxGap, 1)` percentage points because Table S2 counts
per-subject maximum-posterior class assignments (which depend on each patient's
observed data) rather than the prior mixture weights. Table S2's footnote
describes its percentages as "calculated based on the parameter estimation of P1
and P2", which reads as though the two should coincide; they do not, and the
prior probabilities are what the packaged `tp1`/`tp2` reproduce. Table S2 also
resolves the per-arm split that the single pair `TP1`/`TP2` cannot express: the
published model has one set of mixture weights for both dose arms, so the
packaged model cannot by itself reproduce the arm-specific membership shift that
is the paper's main dose finding.

**Baseline in the no-growth class is `TB0 + TBss`, not `TB0`.** This follows
directly from the published equation `TB(t) = TB0*exp(-TS*t) + TBss` evaluated
at *t* = 0, and is noted because "TB0" is labelled "baseline tumor burden" in
Table 2, which is accurate only for the fast and intermediate classes.

**Simulation assumptions made here, not in the paper.**

* The `LDHR` distribution is assumed log-normal with median exactly 1.0
  (Feng 2019 Table 1) and `sdlog = 0.6`, truncated to the reported 0.4-40.5
  range. The paper reports only the median and range, so the shape is a choice.
* `CAV` is applied per dose arm at its typical value, derived from
  `Feng_2014_ipilimumab` with random effects zeroed at the population mean body
  weight of 79.9 kg and the Feng 2014 reference LDH of 206 U/L. Feng 2019 used
  patient-specific empirical-Bayes Cavg1 values, so the cohort here carries no
  between-subject exposure variability.
* The Cavg1 averaging window is taken as the first 21-day (q3w) dosing interval.
  Feng 2019 says only "time-averaged concentration after the first dose". The
  steady-state-equivalent reading `Dose/(CL*tau)` gives 31.7 and 105.8 ug/mL
  instead of 20.6 and 68.6; the conclusions above are unchanged either way.
* Mixture class membership is drawn independently of dose arm, because the
  published model has a single set of mixture weights. The real study showed
  arm-dependent membership (Table S2).

**Not packaged from this paper.** The nonmixture TGD model (TGD-Model 4) and the
two-subpopulation mixture (TGD-Models 2, 8-10) are model-development
intermediates superseded by TGD-Model 11, and are recorded only as BIC rows in
Table S1 without parameter estimates.

# Session info

```{r session}
sessionInfo()
```
