---
title: "FiedlerKelly_2020_fremanezumab_em"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FiedlerKelly_2020_fremanezumab_em}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
```

```{r setup}
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("FiedlerKelly_2020_fremanezumab_em"))$meta$reference`
* Description: `r rxode2::rxode2(readModelDb("FiedlerKelly_2020_fremanezumab_em"))$meta$description`
* Article: <https://doi.org/10.1111/head.13855>

Fiedler-Kelly 2020 develops two exposure-response (E-R) models for fremanezumab, one in episodic migraine (EM, this vignette) and one in chronic migraine (CM, separate vignette `FiedlerKelly_2020_fremanezumab_cm`). The fremanezumab population PK is *not* fitted in this paper; the per-subject Cav values that drive the E-R were taken from individual empirical-Bayes estimates of the Fiedler-Kelly 2019 popPK model (`Fiedler-Kelly_2019_fremanezumab` in this library).

The EM model relates monthly migraine days to fremanezumab Cav and time-on-treatment via three additive components:

$$\text{migraineDays}(t, C_{av}) = \text{BL}_i - e^{\theta_{P,i}\,t} - \text{BL}_i \cdot E_{\max,i}\,\frac{C_{av}}{EC_{50} + C_{av}}$$

with the individual baseline a piecewise-linear function of baseline acute-medication days,

$$\text{BL}_i = \theta_{BL} + \theta_{S}\,\max(0,\,\text{ACUTE\_MED\_DAYS} - 5) + \eta_{\text{BL},i}$$

(breakpoint at 5 d/mo per the medication-overuse-headache convention). `t` is in months (28-day periods). The placebo time-course follows the operator-confirmed Figure 2A form (exp-decay of reduction); the drug-effect uses the Emax/EC50 of Cav reported in Supplementary Table S3.

# Population

The model was fitted to **4,444 monthly migraine-day observations from 1,142 episodic-migraine patients** pooled across two studies — the LBR-101-022 phase 2b study (n = 286) and the TV48125-CNS-30050 phase 3 study (n = 856). Patients met ICHD-3 criteria for episodic migraine: headache on 6–14 (phase 3) or 8–14 (phase 2b) days per month with at least 4 migraine days/month. Pooled demographics (Supplementary Table S1): mean age 41.8 years (range 18–70); mean baseline weight 73.99 kg (range 43.1–120.0); 85.7% female; 79.9% White, 11.2% Black, 7.0% Asian; 12.3% Hispanic ethnicity; mean years since onset 19.97 (range 0–65); mean baseline migraine days 9.7 (range 3–20); mean baseline acute-medication days 3.50 d/mo (range 0–15.2). Concomitant analgesic use 6.7% and concomitant migraine-preventive use 21.4%. Patients received fremanezumab 225 mg monthly, 675 mg monthly, 675 mg quarterly, 225 mg monthly with a 675 mg starting dose, or placebo SC for 3 months; observation unit is one 28-day month.

The same demographic summary is exposed programmatically via the model metadata:

```{r population}
str(rxode2::rxode2(readModelDb("FiedlerKelly_2020_fremanezumab_em"))$meta$population)
```

# Source trace

Per-parameter origin is recorded as an in-file comment next to each `ini()` entry in `inst/modeldb/specificDrugs/FiedlerKelly_2020_fremanezumab_em.R`. The table below collects the equations and parameters in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Composite endpoint equation `migraineDays = BL - exp(exp_PLC * t) - BL * Emax * Cav/(EC50+Cav)` | n/a | Figure 2A; Methods — E-R Analysis Methodology |
| Piecewise-linear baseline `BL = bl_em + slope_AM * max(0, ACUTE_MED_DAYS - 5)` | n/a | Results — Monthly Migraine Days in Patients With EM |
| `bl_em` (typical baseline at AM ≤ 5 d/mo) | 8.35 d/mo | Table S3 |
| `slope_AM` (slope on AM days > 5) | 0.438 d/d | Table S3 |
| `exp_PLC` (placebo time-course exponent, FIXED) | 0.360 / month | Table S3 |
| `Emax_drug` typical (logit-transformed in `ini()`) | 0.252 (fractional) | Table S3 |
| `EC50_drug` | 3.60 µg/mL | Table S3 (NE for IIV) |
| IIV `bl_em` / `slope_AM` (shared additive eta) | SD 1.61 (variance 2.59) | Table S3 |
| IIV `exp_PLC` (additive eta on exponent) | SD 2.92 (variance 8.53) | Table S3 |
| IIV `Emax_drug` (logit-normal eta) | omega² = 0.335 (43.3 %CV per footnote a) | Table S3 |
| Additive residual SD on monthly migraine days | SD 2.35 (variance 5.52) | Table S3 |

Sanity check at the reference (typical-value) baseline with AM = 0 d/mo and Cav = 0 (placebo):

```{r sanity}
mod <- readModelDb("FiedlerKelly_2020_fremanezumab_em")
ev_check <- data.frame(
  id   = 1L,
  time = 0:3,
  CAV  = 0,
  ACUTE_MED_DAYS = 0
)
sim_check <- rxode2::rxSolve(mod |> rxode2::zeroRe(), events = ev_check, returnType = "data.frame")
sim_check[, c("time", "migraineDays")]
```

The month-3 placebo migraine-day count is `r round(sim_check$migraineDays[sim_check$time == 3], 2)`, a reduction of `r round(8.35 - sim_check$migraineDays[sim_check$time == 3], 2)` days from the typical baseline — matching the paper's narrative "approximately 3 days per month at 3 months" placebo reduction.

# Virtual cohort

The original migraine-day diary data are not publicly available. For validation we construct a virtual EM cohort whose covariate distributions match Supplementary Table S1, and we simulate the four trial regimens (placebo; 225 mg q1m; 675 mg q3m; 225 mg q1m with 675 mg starting dose) using a fixed schedule of period-mean Cav values. Cav is supplied as a per-period covariate column (set to 0 for placebo periods).

```{r cohort}
set.seed(20260427)

n_per_arm <- 250L

# Per-period (28-day) Cav values for each regimen.
# Median Cav values stated in Fiedler-Kelly 2020 Results — Chronic Migraine
# section: 28 ug/mL for the 675 mg q3m regimen, 70 ug/mL for the 225 mg q1m
# with 675 mg starting-dose regimen. The 225 mg q1m without loading-dose
# Cav is approximated at the 28 -> 70 trajectory midpoint.
regimen_cav <- list(
  placebo            = c(0, 0, 0),
  q1m_225            = c(35, 50, 60),    # 225 mg monthly, no LD (build-up)
  q3m_675            = c(70, 28, 28),    # 675 mg quarterly: high in mo 1, lower mo 2-3
  q1m_225_with_LD    = c(70, 70, 70)     # 225 mg q1m + 675 mg starting dose: ~steady high
)

make_arm <- function(arm_name, cav_per_month, n, id_offset) {
  expand.grid(
    id    = id_offset + seq_len(n),
    time  = 1:3
  ) |>
    arrange(id, time) |>
    mutate(
      regimen        = arm_name,
      ACUTE_MED_DAYS = pmax(0, rnorm(n(), mean = 3.50, sd = 4.17)),
      CAV            = cav_per_month[time]
    )
}

events <- bind_rows(
  make_arm("placebo",         regimen_cav$placebo,         n_per_arm, id_offset =      0L),
  make_arm("q1m_225",         regimen_cav$q1m_225,         n_per_arm, id_offset =   1000L),
  make_arm("q3m_675",         regimen_cav$q3m_675,         n_per_arm, id_offset =   2000L),
  make_arm("q1m_225_with_LD", regimen_cav$q1m_225_with_LD, n_per_arm, id_offset =   3000L)
) |>
  mutate(evid = 0)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

We simulate two views: (a) typical-value trajectories with `rxode2::zeroRe()` to reproduce the figure-style mean lines, and (b) a stochastic cohort with full IIV for percentile envelopes.

```{r simulate}
mod <- readModelDb("FiedlerKelly_2020_fremanezumab_em")

sim_typ <- rxode2::rxSolve(
  mod |> rxode2::zeroRe(),
  events     = events,
  keep       = c("regimen"),
  returnType = "data.frame"
)

sim_iiv <- rxode2::rxSolve(
  mod,
  events     = events,
  keep       = c("regimen"),
  returnType = "data.frame"
)
```

# Replicate published figures

Figures 3a-b and 4a of Fiedler-Kelly 2020 plot the mean (± 1 SD) monthly migraine days and percent of responders over the three-month follow-up.

```{r figure-3a}
# Replicates Figure 3A of Fiedler-Kelly 2020: mean monthly migraine days by regimen.
sim_iiv |>
  group_by(regimen, time) |>
  summarise(
    mean_md = mean(migraineDays),
    sd_md   = sd(migraineDays),
    .groups = "drop"
  ) |>
  ggplot(aes(time, mean_md, colour = regimen, fill = regimen)) +
  geom_ribbon(aes(ymin = mean_md - sd_md, ymax = mean_md + sd_md), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  labs(
    x        = "Time (months)",
    y        = "Mean monthly migraine days",
    title    = "Figure 3A — mean (±1 SD) monthly migraine days, virtual EM cohort",
    caption  = "Replicates Figure 3A of Fiedler-Kelly 2020 in the virtual cohort."
  ) +
  theme_minimal()
```

```{r figure-4a}
# Replicates Figure 4A of Fiedler-Kelly 2020: percent responders (>= 50%
# reduction from baseline migraine days) by month and regimen.
baseline_per_id <- events |>
  group_by(id, regimen) |>
  summarise(BL = first(8.35 + 0.438 * pmax(0, ACUTE_MED_DAYS - 5)), .groups = "drop")

responders <- sim_iiv |>
  left_join(baseline_per_id, by = c("id", "regimen")) |>
  mutate(reduction_pct = 100 * (BL - migraineDays) / BL,
         responder     = reduction_pct >= 50)

responder_summary <- responders |>
  group_by(regimen, time) |>
  summarise(pct_responders = mean(responder) * 100, .groups = "drop")

ggplot(responder_summary, aes(time, pct_responders, colour = regimen)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(
    x       = "Time (months)",
    y       = "Percent responders (≥ 50% reduction)",
    title   = "Figure 4A — simulated percent responders by month",
    caption = "Replicates Figure 4A of Fiedler-Kelly 2020."
  ) +
  theme_minimal()
```

# Comparison against published narrative

Fiedler-Kelly 2020 Results — EM section reports specific narrative numbers that should reproduce in the typical-value simulation:

```{r narrative-comparison}
narrative_compare <- sim_typ |>
  filter(time == 3) |>
  group_by(regimen) |>
  summarise(
    typical_md_month3 = round(mean(migraineDays), 2),
    typical_reduction = round(8.35 - mean(migraineDays), 2),
    .groups = "drop"
  )

knitr::kable(
  narrative_compare,
  caption = "Typical-value month-3 migraine days and reduction from baseline by regimen."
)
```

The placebo arm produces a `r round(narrative_compare$typical_reduction[narrative_compare$regimen == "placebo"], 2)`-day month-3 reduction (paper: "approximately 3 days"). The 225 mg q1m + 675 mg LD arm achieves a `r round(narrative_compare$typical_reduction[narrative_compare$regimen == "q1m_225_with_LD"], 2)`-day reduction at month 3. The maximal Cav-driven reduction in the model is `r round(8.35 * 0.252, 2)` days, i.e. ~25% of typical baseline (paper: "approximately 25% additional maximal reduction (ie, approximately 2 days)").

# PKNCA validation

PKNCA is the wrong validation target for this model: there is no concentration profile to integrate (Cav is supplied as a covariate, not derived from a PK ODE), and the response variable is a count of migraine days per month rather than a sampled concentration. The validation strategy adopted here is therefore the *narrative-comparison* table immediately above, mirroring the operator-confirmed Figure 2A interpretation and matching the per-regimen reduction ranges reported in the Results section.

# Assumptions and deviations

- **Time unit is months (28-day periods).** The paper aggregates outcomes in monthly periods and reports placebo and drug-effect parameters with implicit "month" as the time unit. The model's `units$time = "month"` documents this; users supplying time in days will need to divide by 28.
- **Placebo time-course form is operator-confirmed from Figure 2A.** Supplementary Table S3 lists only an `Exponent for placebo time-course`. The functional form `BL - exp(exponent * t)` was visually read from Figure 2A by the operator during sidecar request 3 of this extraction. With the typical exponent 0.360 the model gives a 2.94-day placebo reduction at month 3 (matching the paper's narrative "approximately 3 days").
- **Cav as a per-period covariate, not a model output.** This is a PD-only file. The CAV column must be supplied per row by the user, derived externally from the Fiedler-Kelly 2019 popPK model (`Fiedler-Kelly_2019_fremanezumab`) or from observed exposure data. CAV = 0 in placebo periods.
- **Cav values in the virtual cohort.** The per-period regimen Cav profiles in the cohort builder are simplified hand-tuned values intended to bracket the observed median Cavs reported in the Results section (28 µg/mL for 675 mg q3m, 70 µg/mL for 225 mg q1m + LD). For exact reproduction of the paper's full simulation, drive the model with EB Cav from the Fiedler-Kelly 2019 popPK simulator.
- **`ACUTE_MED_DAYS` distribution.** The virtual cohort draws ACUTE_MED_DAYS from N(3.50, 4.17) truncated at 0 (Supplementary Table S1 mean and SD across the pooled EM cohort). The actual data have a heavily right-skewed distribution with median 0.97, so the typical-value simulation slightly over-represents the high-AM tail.
- **Observation variable is `migraineDays`, not `Cc`.** The `checkModelConventions()` warning recommending rename to `Cc` is not appropriate for a count-of-days endpoint that is not a concentration. Following the same convention as `Mulyukov_2018_ranibizumab` (which uses `bcva` as the output).
- **`units$dosing` carries an explanatory string** to satisfy the `checkModelConventions()` requirement; this PD-only model does not consume dose events. The model's input is the CAV covariate column.
- **MU-referencing warning.** `nlmixr2`'s parser flags `etalogitEmax` (and the analogous CM-side `etalogitDrugInt`) as "non-mu-referenced" because the link from the typical theta to the individual parameter goes through an inverse-logit transform rather than a simple addition. This affects estimation behaviour (the model is intended for simulation, not refitting) and is documented here for readers who attempt to refit the model on new data.
