---
title: "Ivermectin (Fimbo 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ivermectin (Fimbo 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Fimbo_2023_ivermectin"))
cat("* Citation: ", ui$reference, "\n\n", sep = "")
cat("* Description: ", ui$description, "\n\n", sep = "")
```

* Article: <https://doi.org/10.1002/psp4.13038>
* Supplement: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10725270/>
  Data S1 holds Figures S1-S3, Data S2 holds Tables S1-S3, and Data S3 holds
  the final NONMEM control stream. Note that the paper's Methods misdirects
  here -- it says "model code is presented in Data S1", but the model code is
  in Data S3; Data S1 contains only the supplementary figures.

Fimbo 2023 fitted a two-compartment model with Savic transit absorption to
1404 ivermectin plasma samples from 468 people dosed during a lymphatic
filariasis mass drug administration (MDA) campaign in the Mkinga district of
Tanga region, Tanzania. Dose was assigned by height pole (3, 6, 9, or 12 mg),
which the paper's second contribution shows systematically underdoses heavier
adults relative to weight-based dosing.

# Population

```{r population, echo = FALSE}
pop <- ui$population
tibble::tibble(Field = names(pop),
               Value = vapply(pop, \(x) paste(
                 if (is.null(names(x))) x else paste0(names(x), " ", x),
                 collapse = "; "), character(1))) |>
  knitr::kable(caption = "Population metadata carried by the model file.")
```

A total of 468 MDA-eligible individuals aged 5-78 years and weighing more than
15 kg were enrolled (Fimbo 2023 Table 1). Sex was recorded for 466: 279 male
(59.9%) and 187 female (40.1%). Because dose is assigned by height pole, the
dose pools are also weight and age strata, and the two are strongly confounded
by construction:

```{r cohort-table, echo = FALSE}
dose_pools <- tibble::tribble(
  ~pool,   ~dose_mg, ~n,  ~wt_mean, ~wt_sd, ~age_mean, ~height_cm,
  "3 mg",         3,  12,    19.92,   3.28,      6.42,   "90-119",
  "6 mg",         6,  49,    25.32,   3.97,     10.92,  "119-139",
  "9 mg",         9, 159,    52.26,  15.17,     36.77,  "139-159",
  "12 mg",       12, 248,    61.73,  10.79,     39.98,     ">159"
)
dose_pools |>
  dplyr::rename(
    "Dose pool"        = pool,
    "Dose (mg)"        = dose_mg,
    "n"                = n,
    "Weight mean (kg)" = wt_mean,
    "Weight SD (kg)"   = wt_sd,
    "Age mean (years)" = age_mean,
    "Height pole (cm)" = height_cm
  ) |>
  knitr::kable(caption = "Fimbo 2023 Table 1, per dose pool. Doses were assigned by the height pole shown in the last column.")
```

The genotype panel (ABCB1 c.3435C>T, ABCB1 rs3842, CYP2C19, CYP2C9, CYP2J2,
CYP3A4 *1B, CYP3A5) is in Fimbo 2023 Table S2; none of these was retained
(Table S3 runs 16-19). They are recorded in the model file's
`covariatesDataExcluded` list for provenance.

# Source trace

Every value below is also carried as an in-file comment beside its `ini()`
entry in `inst/modeldb/specificDrugs/Fimbo_2023_ivermectin.R`.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL/F at 70 kg) | 7.698 L/h | Table 2, row "CL (L/h)", RSE 7% |
| `lvc` (Vc/F at 70 kg) | 146.1 L | Table 2, row "Vc (L)", RSE 11% |
| `lq` (Q/F) | 20.42 L/h | Table 2, row "Q (L/h)", RSE 12% |
| `lvp` (Vp/F) | 207.1 L | Table 2, row "Vp (L)", RSE 11% |
| `lmtt` (MTT) | 1.523 h | Table 2, row "MTT (h)", RSE 5% |
| `lka` (ka) | 0.7082 /h | Table 2, row "K a (/h)", RSE 17% |
| `nn_fix` (NN) | 6 (fixed) | Table 2, row "NN" = "6 (Fixed to this value)"; Data S3 `$THETA (6) FIX` |
| `lfdepot` (reference F) | 1 (fixed) | Results: "F1 was fixed to 1"; Data S3 `TVBIO = 1 * (1 + THETA(10))**DOSE3MG` |
| `e_wt_cl` | 0.75 (fixed) | Results equation `CLi = CLpop x (WT/70)^0.75`; Data S3 `TVCL = THETA(1) * (WT/70)**0.75` |
| `e_wt_vc` | 1 (fixed) | Results equation `Vci = Vcpop x (WT/70)`; Data S3 `TVV2 = THETA(2) * (WT/70)` |
| `e_dose_3mg_fdepot` | 0.4822 | Table 2, "Proportional increase in bioavailability for the 3 mg dose", RSE 40% |
| `e_mix_long_mtt_mtt` | 0.9695 | Table 2, "Proportional difference in MTT between the subpopulation and the rest", RSE 12% |
| Mixture fraction P(long MTT) | 0.1606 | Table 2, "Proportion of a subpopulation with a different typical MTT", RSE 30% (carried in `covariateData`, not `ini()`, because it is a simulation input rather than a model parameter) |
| `etalcl` | 0.4458^2 | Table 2, "Interindividual variability for CL (%CV)" |
| `etalvc` | 0.3339^2 | Table 2, "Interindividual variability for Vc (%CV)" |
| `etalq` | 0.4856^2 | Table 2, "Interindividual variability for Q (%CV)" |
| `etalvp` | 0.4125^2 | Table 2, "Interindividual variability for Vp (%CV)" |
| `etalmtt` | 0.2589^2 | Table 2, "Interindividual variability for MTT (%CV)" |
| `etalfdepot` | 0.278^2 | Table 2, "Interindividual variability for F (%CV)" |
| IIV on ka | none | Data S3 `$OMEGA BLOCK(1) 0 FIX ; EKA` |
| `propSd` | 0.262 | Table 2, "Proportional residual error", RSE 3% |
| Additive residual error | absent | Data S3 `$THETA (0) FIX ; ADD` |
| `d/dt(depot) <- transit(nn, mtt, fdepot) - ka * depot` | n/a | Data S3 `$DES` line 1 (analytical Savic chain with `KTR = (NN+1)/MTT`, `LNFAC = log(6!)`) |
| `f(depot) <- 0` | n/a | Data S3 `$PK` line `F1 = 0`, commented ";;Prevent NONMEM from administering dose into absorption compartment as this will be done through transit compartments" |
| `Cc <- 1000 * central / vc` | n/a | Data S3 `S2 = V2/1000 ; dose in mg, DV in ng/mL` |

## Two readings the supplement settled

Two values in Table 2 are ambiguous from the main text alone, and the Data S3
control stream resolves both. Recording them here because getting either wrong
changes the model materially.

**The IIV column is `omega`, not `%CV`.** Table 2 labels its variability rows
"Interindividual variability for CL (%CV)" with the footnote
"Inter-individual variabilities are on a standard deviation scale (%CV)", while
the Methods define `CV% = 100 * sqrt(exp(omega^2) - 1)`. Those two readings
differ (for CL, `omega = 0.4458` versus `CV = 0.4689`). Data S3 decides it: the
control stream's `$OMEGA` initial estimates are exactly the **squares** of the
corresponding Table S3 `run10` entries.

```{r omega-adjudication, echo = FALSE}
tibble::tibble(
  Parameter    = c("CL", "Vc", "Q", "Vp", "MTT", "F"),
  `Data S3 $OMEGA` = c(0.201, 0.0758, 0.247, 0.172, 0.16, 0.0922),
  `Table S3 run10` = c(0.4479, 0.2753, 0.4971, 0.4147, 0.4002, 0.3036)
) |>
  dplyr::mutate(`sqrt($OMEGA)` = round(sqrt(`Data S3 $OMEGA`), 4)) |>
  knitr::kable(caption = "The reported variability values are the square roots of the NONMEM OMEGA variances, i.e. omega on the eta scale -- not the CV. The model file therefore squares the Table 2 values.")
```

**The residual error is purely proportional.** Table 2 lists only a
proportional term, and Data S3 confirms the additive term was fixed out:
`$THETA (0) FIX ; ADD`, entering `W = SQRT(ADD**2 + PROP**2*IPRED**2)`.

# Structural checks

Two closed-form identities that must hold for a linear two-compartment model,
independent of any published number. These compare the solver against algebra,
so the difference is pure numerical error and a tight bound is appropriate.

```{r analytic-helpers}
mod <- readModelDb("Fimbo_2023_ivermectin")

# Linear-trapezoidal AUC over whatever grid is supplied. Sorts defensively so
# the result never depends on incoming row order.
trapz <- function(time, conc) {
  o <- order(time); time <- time[o]; conc <- conc[o]
  sum(diff(time) * (head(conc, -1) + tail(conc, -1)) / 2)
}

# Analytic terminal (beta) half-life of a 2-compartment model.
beta_half_life <- function(cl, vc, q, vp) {
  kel <- cl / vc; k12 <- q / vc; k21 <- q / vp
  b <- kel + k12 + k21
  beta <- 0.5 * (b - sqrt(b^2 - 4 * kel * k21))
  log(2) / beta
}

# Typical (population) parameters at a given weight, per the model equations.
typ_par <- function(wt) {
  list(cl = 7.698 * (wt / 70)^0.75,
       vc = 146.1 * (wt / 70),
       q  = 20.42,
       vp = 207.1)
}
```

```{r typical-sim}
# One typical individual per dose pool, at that pool's mean body weight,
# assigned to the reference (majority) MTT subpopulation. zeroRe() removes all
# IIV, so this simulation is fully deterministic and reproducible in CI.
grid_t <- c(seq(0, 12, by = 0.1), seq(12.5, 72, by = 0.5))

ev_typ <- dplyr::bind_rows(lapply(seq_len(nrow(dose_pools)), function(i) {
  p <- dose_pools[i, ]
  dose_row <- tibble::tibble(
    id = i, time = 0, amt = p$dose_mg, evid = 1L, cmt = "depot"
  )
  obs_rows <- tibble::tibble(
    id = i, time = grid_t, amt = NA_real_, evid = 0L, cmt = "central"
  )
  dplyr::bind_rows(dose_row, obs_rows) |>
    dplyr::mutate(
      pool         = p$pool,
      WT           = p$wt_mean,
      DOSE_3MG     = as.integer(p$dose_mg == 3),
      MIX_LONG_MTT = 0L
    )
})) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(!anyDuplicated(unique(ev_typ[, c("id", "time", "evid")])))

sim_typ <- rxode2::rxSolve(
  rxode2::zeroRe(mod), events = ev_typ,
  keep = c("pool", "WT", "DOSE_3MG")
) |>
  as.data.frame()

stopifnot(all(sim_typ$Cc >= 0), !anyNA(sim_typ$Cc))
```

```{r identity-checks}
# PKNCA on the deterministic typical-value profiles.
nca_typ_in <- sim_typ |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, pool)

nca_typ_in <- dplyr::bind_rows(
  nca_typ_in,
  nca_typ_in |> dplyr::distinct(id, pool) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, pool, time, .keep_all = TRUE) |>
  dplyr::arrange(id, pool, time)

dose_typ <- ev_typ |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, pool)

nca_typ <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(nca_typ_in, Cc ~ time | pool + id),
  PKNCA::PKNCAdose(dose_typ, amt ~ time | pool + id, route = "extravascular"),
  intervals = data.frame(start = 0, end = Inf,
                         cmax = TRUE, tmax = TRUE,
                         auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE)
))

typ_wide <- as.data.frame(nca_typ) |>
  dplyr::select(pool, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

# Identity 1: AUC(0-inf) must equal 1000 * F * Dose / CL exactly, because the
# model is linear. F = 1.4822 in the 3 mg pool and 1 elsewhere.
# Identity 2: PKNCA's terminal half-life must equal the analytic beta half-life.
ident <- dose_pools |>
  dplyr::left_join(typ_wide, by = "pool") |>
  dplyr::rowwise() |>
  dplyr::mutate(
    fbio        = if (dose_mg == 3) 1.4822 else 1,
    cl_i        = typ_par(wt_mean)$cl,
    auc_closed  = 1000 * fbio * dose_mg / cl_i,
    thalf_closed = do.call(beta_half_life, typ_par(wt_mean)),
    auc_pct     = 100 * (aucinf.obs - auc_closed) / auc_closed,
    thalf_pct   = 100 * (half.life - thalf_closed) / thalf_closed
  ) |>
  dplyr::ungroup()

ident |>
  dplyr::transmute(
    "Dose pool"                 = pool,
    "AUC0-inf, PKNCA"           = round(aucinf.obs, 1),
    "AUC0-inf, F*Dose/CL"       = round(auc_closed, 1),
    "AUC diff (%)"              = round(auc_pct, 3),
    "t1/2, PKNCA"               = round(half.life, 2),
    "t1/2, analytic beta"       = round(thalf_closed, 2),
    "t1/2 diff (%)"             = round(thalf_pct, 3)
  ) |>
  knitr::kable(caption = "Solver vs closed form. Both differences are numerical error only.")

# Deterministic quantities: a tight bound is correct here and will go red on
# any structural mis-transcription (a wrong volume, exponent or unit factor).
stopifnot(
  max(abs(ident$auc_pct))   < 1.0,
  max(abs(ident$thalf_pct)) < 2.0
)
```

The `AUC0-inf = 1000 * F * Dose / CL` identity passing at this tolerance is a
strong joint check on the allometric exponent on CL/F, the 3 mg bioavailability
term, the `f(depot) <- 0` suppression (if the dose bolus were not suppressed,
the depot would receive the dose twice and AUC would double), and the
`1000 *` mg/L to ng/mL conversion.

# Comparison against published NCA

Fimbo 2023 Table 3 reports observed and model-estimated secondary PK parameters
per dose pool. The model-estimated column is the comparison target; it is the
column that also carries `AUC0-inf` and terminal half-life, which cannot be
computed from the observed data (sampling stopped at 6 h).

```{r nca-compare}
published <- tibble::tribble(
  ~pool,  ~cmax, ~tmax, ~auclast, ~aucinf.obs, ~half.life,
  "3 mg",    41,   3.5,      147,        1534,         66,
  "6 mg",    50,   4.0,      171,        1770,         58,
  "9 mg",    49,   4.5,      169,        1521,         41,
  "12 mg",   54,   4.5,      190,        1681,         38
)

# Table 3's AUC0-6 is a 0-6 h partial area; recompute it on that window rather
# than reusing auclast (which runs to 72 h here).
nca_06 <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(nca_typ_in, Cc ~ time | pool + id),
  PKNCA::PKNCAdose(dose_typ, amt ~ time | pool + id, route = "extravascular"),
  intervals = data.frame(start = 0, end = 6, auclast = TRUE)
))

simulated <- typ_wide |>
  dplyr::select(pool, cmax, tmax, aucinf.obs, half.life) |>
  dplyr::left_join(
    as.data.frame(nca_06) |>
      dplyr::filter(PPTESTCD == "auclast") |>
      dplyr::select(pool, auclast = PPORRES),
    by = "pool"
  )

cmp <- simulated |>
  tidyr::pivot_longer(-pool, names_to = "PPTESTCD", values_to = "Simulated") |>
  dplyr::left_join(
    published |> tidyr::pivot_longer(-pool, names_to = "PPTESTCD",
                                     values_to = "Published"),
    by = c("pool", "PPTESTCD")
  ) |>
  dplyr::mutate(
    pct_diff = 100 * (Simulated - Published) / Published,
    label = dplyr::recode(PPTESTCD,
      cmax = "Cmax (ng/mL)", tmax = "Tmax (h)",
      auclast = "AUC0-6 (h*ng/mL)", aucinf.obs = "AUC0-inf (h*ng/mL)",
      half.life = "t1/2 (h)")
  )

cmp |>
  dplyr::transmute(
    "Dose pool"     = pool,
    "NCA parameter" = label,
    "Simulated"     = round(Simulated, 1),
    "Published"     = Published,
    "Difference (%)" = round(pct_diff, 1)
  ) |>
  knitr::kable(caption = "Typical-value simulation vs Fimbo 2023 Table 3 (model-estimated column).")
```

```{r nca-gate}
# Everything compared here is deterministic (zeroRe + fixed weights), so these
# bounds do not depend on the solver thread count. They are set from the
# structural argument below, not from one observed run.
#
# Two systematic reasons the agreement is not exact, both one-directional and
# both understood:
#   1. Table 3's model-estimated values are means over the paper's individual
#      (EBE-based) estimates, whereas this is a typical-value prediction at the
#      pool's MEAN weight. For a log-normal eta the two differ.
#   2. Table 1 reports mean weight; the typical-value prediction is really a
#      median-weight quantity. The pools are not perfectly symmetric in weight.
stopifnot(
  # Exposure and peak: a mis-transcribed clearance, volume, dose or unit factor
  # moves these by tens of percent.
  max(abs(cmp$pct_diff[cmp$PPTESTCD %in% c("aucinf.obs", "auclast", "cmax")])) < 20,
  # Half-life is a pure disposition quantity and reproduces much more tightly.
  max(abs(cmp$pct_diff[cmp$PPTESTCD == "half.life"])) < 10,
  # Tmax is on a 0.1 h grid up to 12 h; the published values are 3.5-4.5 h.
  max(abs(cmp$Simulated[cmp$PPTESTCD == "tmax"] -
            cmp$Published[cmp$PPTESTCD == "tmax"])) < 1.5
)
```

The half-life column is the most informative row of that table. Fimbo 2023
reports terminal half-life falling from 66 h in the 3 mg pool to 38 h in the
12 mg pool and attributes it (Discussion) to "the covariate effect of weight on
CL/F and Vc/F". The packaged model reproduces that gradient because allometry
is applied to CL/F and Vc/F but **not** to Q/F or Vp/F: as weight rises, the
central volume grows while the peripheral volume stays at 207 L, so the
peripheral compartment becomes relatively smaller and the terminal phase
shortens. Encoding allometry on all four disposition parameters -- the more
common convention, and what Fimbo 2023's own Figure S1 calls "Duthaler's 2020
model" -- would flatten this gradient and lose the result.

# Virtual cohort and visual predictive check

```{r cohort}
# rxode2's RNG streams are partitioned per solver thread, so this cohort is not
# byte-identical across machines with different thread counts. Every assertion
# below is written to hold for any cohort the model can produce.
set.seed(20231215)

n_per_pool <- 200L   # cap is 200/arm; the published pools are 12/49/159/248

make_pool <- function(i) {
  p <- dose_pools[i, ]
  wt <- rnorm(n_per_pool, p$wt_mean, p$wt_sd)
  # The MDA programme excludes anyone under 15 kg, so truncate rather than
  # allowing the normal tail to produce implausible weights.
  wt <- pmin(pmax(wt, 15), 95)
  subj <- tibble::tibble(
    id           = (i - 1L) * n_per_pool + seq_len(n_per_pool),
    WT           = wt,
    DOSE_3MG     = as.integer(p$dose_mg == 3),
    MIX_LONG_MTT = rbinom(n_per_pool, 1L, 0.1606),
    pool         = p$pool,
    dose_mg      = p$dose_mg
  )
  dplyr::bind_rows(
    subj |> dplyr::mutate(time = 0, amt = dose_mg, evid = 1L, cmt = "depot"),
    subj |> tidyr::crossing(time = grid_t) |>
      dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
  ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(lapply(seq_len(nrow(dose_pools)), make_pool))
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))

sim <- rxode2::rxSolve(
  mod, events = events, keep = c("pool", "WT", "dose_mg", "MIX_LONG_MTT")
) |>
  as.data.frame() |>
  dplyr::mutate(pool = factor(pool, levels = dose_pools$pool))
```

```{r figure-1, fig.cap = "Replicates the structure of Figure 1 of Fimbo 2023: simulated 5th, 50th and 95th percentiles of ivermectin concentration by dose pool over the 0-6 h sampling window used in the study."}
sim |>
  dplyr::filter(time <= 6) |>
  dplyr::group_by(pool, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue4", linewidth = 0.8) +
  facet_wrap(~pool, nrow = 1) +
  labs(x = "Time (h)", y = "Ivermectin concentration (ng/mL)",
       caption = "Replicates Figure 1 of Fimbo 2023 (sampling window only).") +
  theme_bw()
```

```{r vpc-check}
# The observed AUC0-6 and Cmax in Table 3 sit inside the simulated cohort's
# central range. Compare the cohort MEDIAN, which is robust to the tails and,
# for a log-normal eta, is the quantity the typical-value prediction estimates.
obs_tab3 <- tibble::tribble(
  ~pool,   ~cmax_obs,
  "3 mg",         40,
  "6 mg",         52,
  "9 mg",         51,
  "12 mg",        57
)

cmax_sim <- sim |>
  dplyr::filter(time <= 6) |>
  dplyr::group_by(pool, id) |>
  dplyr::summarise(cmax = max(Cc), .groups = "drop") |>
  dplyr::group_by(pool) |>
  dplyr::summarise(cmax_med = median(cmax), .groups = "drop") |>
  dplyr::left_join(obs_tab3, by = "pool") |>
  dplyr::mutate(pct = 100 * (cmax_med - cmax_obs) / cmax_obs)

cmax_sim |>
  dplyr::transmute("Dose pool" = pool,
                   "Simulated median Cmax (ng/mL)" = round(cmax_med, 1),
                   "Observed Cmax, Table 3 (ng/mL)" = cmax_obs,
                   "Difference (%)" = round(pct, 1)) |>
  knitr::kable(caption = "Cohort median Cmax over 0-6 h vs the observed column of Fimbo 2023 Table 3.")

# Robust, cohort-derived: assert on the centre and an envelope, never on an
# extreme. A mis-transcribed volume or dose moves the whole distribution.
stopifnot(
  abs(median(cmax_sim$pct)) < 20,
  max(abs(cmax_sim$pct))    < 30
)
```

```{r mixture-check}
# The MTT mixture shifts absorption timing but must NOT change total exposure,
# because MTT only reshapes the input function -- it does not scale its area.
# This is a structural property of the transit chain and holds per cohort.
by_mix <- sim |>
  dplyr::filter(pool == "12 mg") |>
  dplyr::group_by(MIX_LONG_MTT, id) |>
  dplyr::summarise(auc72 = trapz(time, Cc), tmax = time[which.max(Cc)],
                   .groups = "drop") |>
  dplyr::group_by(MIX_LONG_MTT) |>
  dplyr::summarise(auc72 = median(auc72), tmax = median(tmax), .groups = "drop")

knitr::kable(by_mix, digits = 1,
             caption = "Long-MTT subpopulation (1) vs reference (0), 12 mg pool: later peak, same area.")

auc_ratio <- by_mix$auc72[by_mix$MIX_LONG_MTT == 1] /
  by_mix$auc72[by_mix$MIX_LONG_MTT == 0]
stopifnot(
  # Areas agree; the 10% band admits cohort sampling noise between the two
  # subgroups (the long-MTT group is only ~16% of the cohort).
  abs(auc_ratio - 1) < 0.10,
  # The long-MTT group peaks later. MTT nearly doubles (x1.9695), so this is a
  # large, reliably-signed effect, not a near-zero one.
  by_mix$tmax[by_mix$MIX_LONG_MTT == 1] > by_mix$tmax[by_mix$MIX_LONG_MTT == 0]
)
```

# Height-based versus weight-based dosing

Fimbo 2023's second conclusion is that height-pole dosing underdoses heavier
adults: the top pole caps the dose at 12 mg for everyone taller than 159 cm, so
mg/kg falls as weight rises, whereas the WHO 200 ug/kg target rounded to whole
3 mg tablets tracks weight. The paper reports that 23%, 30%, and 100% of people
weighing 61-70, 71-80, and over 80 kg were underdosed (below 150 ug/kg) under
height-based dosing, versus 0% in every one of those bands under weight-based
dosing (Table S1).

The comparison below is restricted to the weight bands from 51 kg upward.
That restriction matters: the 12 mg dose is the *top* height pole, given to
everyone taller than 159 cm, so it is only the applicable height-based dose for
adults. A 35 kg person is short enough to fall on the 6 or 9 mg pole, and
assigning them 12 mg would be counterfactual -- it would manufacture a large
apparent over-exposure at low weight that the height-pole system never
produces. Because the model carries weight and not height, and no height-weight
relationship is published for this cohort, the honest comparison is over the
weight range where the top pole genuinely applies. This is also exactly the
range the paper's claim is about (above versus below 70 kg).

```{r dosing-comparison}
bands <- tibble::tribble(
  ~band,      ~wt,
  "51-60 kg",  55,
  "61-70 kg",  65,
  "71-80 kg",  75,
  ">80 kg",    85
)

# Weight-based: 200 ug/kg rounded to the nearest whole 3 mg tablet, exactly as
# Fimbo 2023 describes in "Evaluation of ivermectin height-based dosing".
bands <- bands |>
  dplyr::mutate(
    dose_height = 12,                                   # top height pole cap
    dose_weight = pmax(3, round(0.2 * wt / 3) * 3),     # 200 ug/kg -> whole tablets
    ugkg_height = 1000 * dose_height / wt,
    ugkg_weight = 1000 * dose_weight / wt
  )

ev_band <- dplyr::bind_rows(lapply(seq_len(nrow(bands)), function(i) {
  b <- bands[i, ]
  dplyr::bind_rows(lapply(c("Height-based", "Weight-based"), function(strat) {
    amt <- if (strat == "Height-based") b$dose_height else b$dose_weight
    idx <- i + ifelse(strat == "Height-based", 0L, 100L)
    dplyr::bind_rows(
      tibble::tibble(id = idx, time = 0, amt = amt, evid = 1L, cmt = "depot"),
      tibble::tibble(id = idx, time = grid_t, amt = NA_real_, evid = 0L,
                     cmt = "central")
    ) |>
      dplyr::mutate(band = b$band, strategy = strat, WT = b$wt,
                    DOSE_3MG = 0L, MIX_LONG_MTT = 0L)
  }))
})) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(!anyDuplicated(unique(ev_band[, c("id", "time", "evid")])))

sim_band <- rxode2::rxSolve(rxode2::zeroRe(mod), events = ev_band,
                            keep = c("band", "strategy", "WT")) |>
  as.data.frame()

auc_band <- sim_band |>
  dplyr::group_by(band, strategy, WT) |>
  dplyr::summarise(auc = trapz(time, Cc), .groups = "drop") |>
  dplyr::mutate(band = factor(band, levels = bands$band))
```

```{r figure-2, fig.cap = "Replicates the message of Figure 2 of Fimbo 2023: typical-value AUC(0-72 h) by weight band under the 12 mg height-pole cap versus 200 ug/kg rounded to whole 3 mg tablets."}
ggplot(auc_band, aes(band, auc, colour = strategy, group = strategy)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.4) +
  labs(x = NULL, y = "AUC(0-72 h) (h*ng/mL)", colour = NULL,
       caption = "Replicates Figure 2 of Fimbo 2023.") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "top")
```

```{r dosing-gate}
h <- auc_band |> dplyr::filter(strategy == "Height-based") |>
  dplyr::arrange(WT)
w <- auc_band |> dplyr::filter(strategy == "Weight-based") |>
  dplyr::arrange(WT)

# Deterministic (zeroRe, fixed band weights), so these are exact bounds.
#
# The gate encodes the paper's actual claim, which is DIRECTIONAL, not a
# flatness claim: "contrary to weight-based dosing, height-based dosing results
# in relatively lower AUC0-inf for individuals weighing greater than 70 kg
# compared to those weighing less than 70 kg" (Fimbo 2023 Results, Figure 2).
#
# Note it would be wrong to assert that weight-based dosing gives a FLATTER
# exposure profile than height-based dosing. It does not, and the model says so:
# rounding 200 ug/kg to whole 3 mg tablets makes the realised mg/kg jump around
# (171-218 ug/kg across these bands), so the weight-based spread is comparable
# to the height-based one. What weight-based dosing does is keep everyone above
# the 150 ug/kg efficacy threshold; it does not equalise exposure.
mean_gt70 <- function(d) mean(d$auc[d$WT > 70])
mean_le70 <- function(d) mean(d$auc[d$WT <= 70])

stopifnot(
  # Under the fixed 12 mg cap, AUC must fall monotonically with weight, because
  # CL/F rises allometrically while the dose does not rise at all. This is exact
  # and structural -- it goes red on a wrong allometric exponent or reference.
  all(diff(h$auc) < 0),
  # The paper's claim, height-based arm: heavier individuals get LESS exposure.
  mean_gt70(h) < mean_le70(h),
  # The paper's claim, weight-based arm: the pattern is reversed ("contrary to
  # weight-based dosing"), so heavier individuals do NOT lose exposure.
  mean_gt70(w) > mean_le70(w),
  # The paper's arithmetic claim: everyone above 80 kg is underdosed
  # (< 150 ug/kg) by the 12 mg cap, and nobody is under weight-based dosing.
  bands$ugkg_height[bands$band == ">80 kg"] < 150,
  all(bands$ugkg_weight >= 150)
)

bands |>
  dplyr::transmute("Weight band" = band,
                   "Height-based dose (mg)" = dose_height,
                   "Height-based (ug/kg)" = round(ugkg_height),
                   "Weight-based dose (mg)" = dose_weight,
                   "Weight-based (ug/kg)" = round(ugkg_weight)) |>
  knitr::kable(caption = "Dose per kg by weight band under the two strategies. 150 ug/kg is the efficacy threshold Fimbo 2023 uses.")
```

# Assumptions and deviations

- **Cohort size.** The virtual cohort uses 200 subjects per dose pool rather
  than the published 12 / 49 / 159 / 248, to keep Monte-Carlo noise out of the
  cohort-level assertions while respecting the 200-per-arm cap. The
  typical-value comparisons against Table 3 do not depend on cohort size at all.

- **Weight distributions** are drawn as `Normal(mean, SD)` per dose pool from
  Fimbo 2023 Table 1 and truncated to [15, 95] kg. The paper reports only mean
  and SD, not the distributional shape; the lower truncation reflects the
  programme's stated 15 kg eligibility floor. Truncation shifts the 3 mg pool's
  realised mean slightly upward (its mean of 19.92 kg is only 1.5 SD above the
  floor); the 9 and 12 mg pools are essentially unaffected.

- **The comparison target is a typical-value prediction, the published column
  is a mean over individual estimates.** Fimbo 2023 Table 3's "model estimated"
  values are summaries of per-subject empirical-Bayes estimates from the
  original dataset. With only three post-dose samples per subject the etas are
  substantially shrunk, so that mean sits close to a typical-value prediction
  but is not identical to one. This is the main reason the AUC comparison
  agrees to within roughly 5% rather than exactly, and it is why the
  cohort-level checks compare medians rather than means: for a log-normal eta
  the cohort mean of `1/CL` is inflated by `exp(omega^2 / 2)` while the median
  is not.

- **Height is not modelled, so the dosing comparison is restricted to adults.**
  The paper assigns dose by height pole, but the model carries weight, not
  height, and no height-weight relationship is published for this cohort. The
  height-versus-weight section therefore compares the 12 mg top-pole cap
  against weight-based dosing at fixed band weights, rather than reconstructing
  individual pole assignments. It covers only the 51 kg and heavier bands,
  because the 12 mg pole applies to people taller than 159 cm: assigning 12 mg
  to a 35 kg child would be counterfactual and would fabricate an
  over-exposure at low weight that the pole system does not produce.

- **Weight-based dosing is not "flatter", and the vignette does not claim it
  is.** An earlier draft of this vignette asserted that weight-based dosing
  keeps AUC flatter across weight bands than height-based dosing. The model
  falsifies that: rounding the WHO 200 ug/kg target to whole 3 mg tablets makes
  the realised dose intensity swing between 171 and 218 ug/kg, so the
  weight-based AUC spread is comparable to the height-based one. What
  weight-based dosing achieves -- and all the paper claims for it -- is that
  nobody falls below the 150 ug/kg efficacy threshold, and that exposure stops
  declining with weight. The gate now tests that directional claim rather than
  a flatness claim.

- **Mixture assignment.** `MIX_LONG_MTT` is a latent class, not an observable
  patient characteristic. The cohort draws it as `Bernoulli(0.1606)`; the
  typical-value simulations set it to 0 (the 83.9% majority class).

- **IIV correlations are unreported.** Data S3 declares `$OMEGA BLOCK(4)` over
  CL, Vc, Q and Vp with all six off-diagonal *initial estimates* set to 0. A
  NONMEM `BLOCK` without `FIX` estimates its covariances, so the deposited
  stream implies six correlation parameters were estimated -- but neither
  Table 2 nor Table S3 reports a single covariance or correlation, and Table
  S3's variability columns are diagonals only (`ECL`, `EVC`, `EQ`, `EVP`,
  `EMTT`, `EF1`). The estimated covariances are therefore a genuine reporting
  gap. The model file encodes the four IIVs as independent (diagonal) because
  that is the only structure the published numbers support; a user reproducing
  individual-level correlation structure should treat this as a known
  limitation rather than as a claim that the covariances were zero.

- **MTT mixture effect: 97.4% in prose, 96.95% in the table.** Fimbo 2023's
  Results text describes "a subpopulation (16% of the studied population) with
  97.4% higher MTT", but Table 2 and Table S3 `run15` both give the estimate as
  0.9695, i.e. 96.95%. The abstract rounds it to "97%". The model file encodes
  the table value (0.9695), since the parameter table is the authoritative
  source for a fitted estimate and the prose figure appears nowhere else.

- **Age range discrepancy.** Fimbo 2023 Methods states participants were "aged
  between 5- and 78 years old", but the per-weight-band age ranges in Table S1
  extend to 90 and 91 years. The model's `population` metadata records the
  Methods range and flags the discrepancy. Age is not a covariate in the final
  model, so nothing downstream depends on which reading is correct.

- **Albendazole.** Every participant also received albendazole 400 mg, and the
  paper notes albendazole may compete for CYP3A4/5. No albendazole term is in
  the model, so the ivermectin parameters here are conditional on
  co-administration with albendazole.

- **No non-paper-derived parameter values.** Every value in the model file comes
  from Fimbo 2023 Table 2, Table S3, or the Data S3 NONMEM control stream. The
  structural model was adapted from Duthaler 2019 via the NONMEM PRIOR
  subroutine, but all encoded estimates are Fimbo 2023's own; Duthaler's prior
  values (`$THETA (7.67 FIXED) (89.13 FIXED) (19.0 FIXED) (234 FIXED)`) are
  deliberately not used.
