---
title: "Calcitriol-conjugated quantum dots in mice with inflammatory breast cancer (Forder 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calcitriol-conjugated quantum dots in mice with inflammatory breast cancer (Forder 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Forder J, Smith M, Wagner M, Schaefer RJ, Gorky J, van Golen KL,
  Nohe A, Dhurjati P (2019). A physiologically-based pharmacokinetic model for
  targeting calcitriol-conjugated quantum dots to inflammatory breast cancer
  cells. *Clin Transl Sci* 12(6):617-624.
* Article: <https://doi.org/10.1111/cts.12664> (open access, PMC6853145)
* Supplement: Figures S1-S3, Tables S1-S2 and the authors' complete MATLAB
  model code (`CTS-12-617-s006.docx`), from the EuropePMC supplementary-file
  bundle for PMC6853145.

Forder 2019 builds a small flow-limited PBPK model for the distribution of
fluorescent quantum dots (QDs) after one intravenous injection in mice. Three
QDs are modelled: unconjugated control QDs (ConQD), calcitriol-conjugated QDs
(CalQD), and CalQDs that also carry the SM3 anti-MUC1 antibody (SM3 CalQD),
which targets inflammatory breast cancer (IBC) cells. There are three
connectivities (Figure 1): a healthy (tumor-free) mouse, an early-stage IBC
mouse with a tumor inside breast tissue, and a late-stage IBC mouse with
metastatic tumor tissue in parallel with the other organs. The authors fitted
the partition coefficients separately for each of the 3 x 3 combinations, so
the package has nine models:

```{r model-list}
qds <- c(
  ConQD = "controlQuantumDot",
  CalQD = "calcitriolQuantumDot",
  "SM3 CalQD" = "sm3CalcitriolQuantumDot"
)
stages <- c(Healthy = "healthy", "Early-stage" = "earlyIBC", "Late-stage" = "lateIBC")
grid <- expand.grid(qd = names(qds), stage = names(stages), stringsAsFactors = FALSE)
grid$model <- sprintf("Forder_2019_%s_%s_mouse_pbpk", qds[grid$qd], stages[grid$stage])
stopifnot(all(grid$model %in% modeldb$name))
knitr::kable(grid |> rename(QD = qd, Stage = stage, Model = model))
```

# Population

```{r population}
ui_sm3_late <- rxode2::rxode(readModelDb("Forder_2019_sm3CalcitriolQuantumDot_lateIBC_mouse_pbpk"))
str(ui_sm3_late$population)
```

The biodistribution data are from Schaefer 2012 (reference 10 of the paper):
female mice aged 13-16 weeks (about 22.0 g), with or without SUM149 IBC
tumors, imaged by fluorescence 4 days after injection. Only the mean pixel
intensity per organ (Table 2) was used, and there are no early-stage data, so
the early-stage partition coefficients were fitted to the late-stage
intensities. The number of animals is not reported. No between-subject
variability or residual error was estimated, so every model here is a
deterministic typical-value model.

# Source trace

| Item | Source | Notes |
|---|---|---|
| Organ volumes `v_*` | Table 1; supplementary code `vp`, `vk`, ... | Code values (more digits) used, e.g. `v_liver` 1.2078 vs Table 1 1.208 |
| Organ plasma flows `q_*` | Table 1; supplementary code `qk`, `qli`, ... | See the `q_other` note under Errata |
| `v_spleen` enlarged 0.154 mL | Table 1 (parenthesis) | Used only in the Figure 3 / Table S2 section below |
| `kelim` = 1e-6 | Eq. 3; code `k1 = 0.000001` | Sits inside the `1/V_liver` bracket of Eq. 12 |
| `lkp_other` = log(1) | Results, *Partition coefficients*; code `ro = 1` | Assumed, not fitted |
| `lkp_kidney`, `lkp_liver`, `lkp_spleen`, `lkp_lung` | Table 3 | One column per QD x stage |
| `lkp_tumor` | Table 3 | Early- and late-stage models only |
| `lkp_breast` | Table 3 | Early-stage models only |
| Organ ODE `dC/dt = Q/V (C_p - C/P)` | Eq. 6, 9-11, 13, 16 | |
| Liver ODE (spleen outflow + hepatic artery in, elimination) | Eq. 12 | |
| Healthy plasma ODE | Eq. 7-8 | Identical to the code |
| Late-stage plasma ODE | Supplementary code `Tsm3QD_MW()` | Differs from printed Eq. 14; see Errata |
| Early-stage plasma, breast and tumor ODEs | Eq. 17-18; code `Tsm3QD()` | Breast/tumor coupling differs from printed Eq. 19-20; see Errata |
| IV pulse into plasma; 40 nM initial plasma concentration | Methods; code `ode23s(..., [40; 0; ...])` | Dose here: 40 nM x 1.078 mL = 43.12 pmol |
| Fit target: organ intensities at 96 h | Table 2; code `TSoS()` / `nTSoS()` / `Tsm3QDSoS()` | |
| Enlarged-spleen partition coefficients | Table S2 | Section "Enlarged spleen" |

# Simulation set-up

Every model is dosed with the code's 40 nM initial plasma concentration
(43.12 pmol into `plasma`). Observation rows are placed on the `plasma` ODE
state, and all organ concentrations are returned alongside because each state
already holds a concentration.

```{r sim-helpers}
dose_pmol <- 40 * 1.078
times <- sort(unique(c(seq(0, 12, by = 0.02), seq(12, 96, by = 0.25))))

make_events <- function(times) {
  rxode2::et(amt = dose_pmol, cmt = "plasma") |>
    rxode2::et(times, cmt = "plasma")
}

sim_model <- function(name, times, ...) {
  ui <- rxode2::rxode(readModelDb(name))
  out <- rxode2::rxSolve(ui, make_events(times), returnType = "data.frame", ...)
  if (is.null(out$id)) out$id <- 1L
  out$model <- name
  out
}

sims <- lapply(grid$model, sim_model, times = times)
names(sims) <- grid$model
```

# Reproducing the fit: Table 3 partition coefficients against Table 2

Table 3 reports the partition coefficients that minimise the relative squared
error between simulated and observed day-4 (96 h) intensities for kidney,
liver, spleen and lung. Simulating each packaged model to 96 h and comparing
against Table 2 therefore checks the whole transcription (topology, volumes,
flows, partition coefficients, dose) at once. There is no residual error or
between-subject variability, so this is a deterministic check and a tight
tolerance is appropriate.

```{r table2}
table2 <- tibble::tribble(
  ~qd, ~data, ~kidney, ~liver, ~lung, ~spleen,
  "ConQD", "No tumor", 10.40, 16.23, 4.66, 12.73,
  "CalQD", "No tumor", 11.01, 20.38, 3.16, 10.99,
  "SM3 CalQD", "No tumor", 10.47, 17.18, 5.01, 9.87,
  "ConQD", "Late-stage tumor", 12.54, 14.85, 8.22, 26.43,
  "CalQD", "Late-stage tumor", 15.15, 20.15, 7.59, 21.46,
  "SM3 CalQD", "Late-stage tumor", 11.57, 18.12, 7.46, 17.92
) |>
  pivot_longer(c(kidney, liver, lung, spleen), names_to = "organ", values_to = "observed")

day4 <- bind_rows(sims) |>
  filter(abs(time - 96) < 1e-8) |>
  select(model, kidney, liver, lung, spleen) |>
  left_join(grid, by = "model") |>
  pivot_longer(c(kidney, liver, lung, spleen), names_to = "organ", values_to = "simulated") |>
  mutate(data = ifelse(stage == "Healthy", "No tumor", "Late-stage tumor")) |>
  left_join(table2, by = c("qd", "data", "organ")) |>
  mutate(pct_diff = 100 * (simulated - observed) / observed)

stopifnot(nrow(day4) == 36, !anyNA(day4$pct_diff))

day4 |>
  mutate(simulated = signif(simulated, 4), pct_diff = round(pct_diff, 2)) |>
  select(stage, qd, organ, observed, simulated, pct_diff) |>
  rename(
    Stage = stage, QD = qd, Organ = organ,
    "Table 2 intensity" = observed, "Simulated at 96 h" = simulated,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable()
```

```{r table2-gate}
max_err <- day4 |>
  group_by(stage, qd) |>
  summarise(max_abs_pct = max(abs(pct_diff)), .groups = "drop")
knitr::kable(max_err |> rename(Stage = stage, QD = qd, "Max abs difference (%)" = max_abs_pct), digits = 2)

is_sm3_early <- max_err$stage == "Early-stage" & max_err$qd == "SM3 CalQD"
stopifnot(
  # Eight of nine fits reproduce Table 2 to rounding of the printed
  # partition coefficients.
  all(max_err$max_abs_pct[!is_sm3_early] < 1.5),
  # The early-stage SM3 CalQD column does not (see Errata): it over-predicts
  # every organ by about 3 percent.
  max_err$max_abs_pct[is_sm3_early] > 2.5,
  max_err$max_abs_pct[is_sm3_early] < 4
)
```

Eight of the nine models reproduce the fitted intensities to within the
rounding of the printed partition coefficients. The early-stage SM3 CalQD
model over-predicts every organ by about 3%; the reason is shown in the next
section.

# Enlarged spleen (Table S2)

The authors also refitted the tumor models with the spleen volume doubled to
0.154 mL (splenomegaly), keeping the spleen blood flow (Table S2). These fits
are a sensitivity analysis and are not shipped as separate models; they are
reproduced here by overriding `v_spleen` and the partition coefficients in
the packaged early- and late-stage models.

```{r tableS2}
tableS2 <- tibble::tribble(
  ~stage, ~qd, ~kidney, ~liver, ~spleen, ~lung, ~tumor, ~breast,
  "Early-stage", "ConQD", 51.6, 61.1, 109, 33.8, 33.5, 2.8,
  "Early-stage", "CalQD", 113, 151, 160, 56.7, 63.4, 2.9,
  "Early-stage", "SM3 CalQD", 59.1, 92.6, 91.5, 38.1, 97.4, 1.83,
  "Late-stage", "ConQD", 53.8, 63.6, 113, 35.2, 33.5, NA,
  "Late-stage", "CalQD", 103, 136, 145, 51.4, 45.7, NA,
  "Late-stage", "SM3 CalQD", 55.6, 86.7, 86.1, 35.8, 97.5, NA
)

sim_enlarged <- function(i) {
  r <- tableS2[i, ]
  name <- grid$model[grid$qd == r$qd & grid$stage == r$stage]
  kp <- c(kidney = r$kidney, liver = r$liver, spleen = r$spleen, lung = r$lung, tumor = r$tumor)
  if (!is.na(r$breast)) kp <- c(kp, breast = r$breast)
  params <- c(v_spleen = 0.154, stats::setNames(log(kp), paste0("lkp_", names(kp))))
  ui <- rxode2::rxode(readModelDb(name))
  ui <- suppressMessages(do.call(rxode2::ini, c(list(ui), as.list(params))))
  out <- rxode2::rxSolve(ui, make_events(times), returnType = "data.frame")
  out$stage <- r$stage
  out$qd <- r$qd
  out
}
enlarged <- bind_rows(lapply(seq_len(nrow(tableS2)), sim_enlarged))

s2_check <- enlarged |>
  filter(abs(time - 96) < 1e-8) |>
  select(stage, qd, kidney, liver, lung, spleen) |>
  pivot_longer(c(kidney, liver, lung, spleen), names_to = "organ", values_to = "simulated") |>
  mutate(data = "Late-stage tumor") |>
  left_join(table2, by = c("qd", "data", "organ")) |>
  mutate(pct_diff = 100 * (simulated - observed) / observed)

stopifnot(nrow(s2_check) == 24, !anyNA(s2_check$pct_diff))
knitr::kable(
  s2_check |>
    group_by(stage, qd) |>
    summarise(max_abs_pct = max(abs(pct_diff)), .groups = "drop") |>
    rename(Stage = stage, QD = qd, "Max abs difference vs Table 2 (%)" = max_abs_pct),
  digits = 2
)
stopifnot(max(abs(s2_check$pct_diff)) < 1.5)

# The early-stage SM3 CalQD column of Table 3 is identical to its Table S2
# (enlarged-spleen) column; no other Table 3 / Table S2 pair coincides.
sm3_early_t3 <- rxode2::rxode(readModelDb(
  "Forder_2019_sm3CalcitriolQuantumDot_earlyIBC_mouse_pbpk"
))$theta[paste0("lkp_", c("kidney", "liver", "spleen", "lung", "tumor", "breast"))]
sm3_early_s2 <- unlist(tableS2[tableS2$stage == "Early-stage" & tableS2$qd == "SM3 CalQD", 3:8])
stopifnot(isTRUE(all.equal(unname(exp(sm3_early_t3)), unname(sm3_early_s2))))
```

All six enlarged-spleen fits reproduce Table 2 to rounding. The early-stage
SM3 CalQD partition coefficients printed in Table 3 are exactly the Table S2
enlarged-spleen values. With the doubled spleen volume they fit Table 2
exactly. With the regular spleen volume, which is what Table 3 should describe,
they are about 3% off. The regular-spleen fit for this QD therefore seems to
have been replaced by the enlarged-spleen fit when the table was put together.
The packaged model keeps the printed values (see Errata).

# Mass balance

Elimination is negligible (`kelim` = 1e-6), so the total QD mass
`sum(V_i * C_i)` should stay at the 43.12 pmol dose over 4 days.

```{r mass-balance}
vol_of <- function(name) {
  th <- rxode2::rxode(readModelDb(name))$theta
  th[grepl("^v_", names(th))]
}
mass <- bind_rows(lapply(grid$model, function(name) {
  v <- vol_of(name)
  s <- sims[[name]]
  organs <- sub("^v_", "", names(v))
  data.frame(
    model = name, time = s$time,
    mass = as.vector(as.matrix(s[, organs]) %*% v)
  )
})) |>
  left_join(grid, by = "model")

mass_summary <- mass |>
  group_by(stage, qd) |>
  summarise(
    max_mass = max(mass), mass_96h = mass[which.min(abs(time - 96))],
    .groups = "drop"
  ) |>
  mutate(pct_gain_96h = 100 * (mass_96h - dose_pmol) / dose_pmol)
knitr::kable(
  mass_summary |>
    rename(
      Stage = stage, QD = qd, "Max total (pmol)" = max_mass,
      "Total at 96 h (pmol)" = mass_96h, "Gain at 96 h (%)" = pct_gain_96h
    ),
  digits = 3
)

late <- mass_summary$stage == "Late-stage"
stopifnot(
  # Healthy and early-stage topologies conserve mass exactly.
  all(abs(mass_summary$pct_gain_96h[!late]) < 0.01),
  # The late-stage plasma equation of the authors' code (used here) creates
  # a few percent of spurious mass; see Errata.
  all(mass_summary$pct_gain_96h[late] > 0.5),
  all(mass_summary$pct_gain_96h[late] < 4)
)
```

For comparison, here is the late-stage SM3 CalQD model with the printed Eq. 14
plasma equation, which returns `(q_liver + q_spleen) * liver / kp_liver`
exactly as the healthy model does:

```{r printed-eq14}
ui_late <- rxode2::rxode(readModelDb("Forder_2019_sm3CalcitriolQuantumDot_lateIBC_mouse_pbpk"))
ui_eq14 <- rxode2::model(
  ui_late,
  d / dt(plasma) <- (q_kidney * kidney / kp_kidney +
    (q_liver + q_spleen) * liver / kp_liver +
    q_lung * lung / kp_lung +
    q_other * other / kp_other +
    q_tumor * tumor / kp_tumor -
    q_total * plasma) / v_plasma
)
eq14 <- rxode2::rxSolve(ui_eq14, make_events(c(0, 96)), returnType = "data.frame")
eq14_96 <- eq14[eq14$time == 96, c("kidney", "liver", "lung", "spleen")]
obs_sm3_late <- table2$observed[table2$qd == "SM3 CalQD" & table2$data == "Late-stage tumor"]
eq14_pct <- 100 * (unlist(eq14_96) - obs_sm3_late) / obs_sm3_late
round(eq14_pct, 2)
stopifnot(all(eq14_pct < -3), all(eq14_pct > -4.5))
```

The mass-conserving printed equation under-predicts all four Table 2
intensities by about 3.5% with the Table 3 partition coefficients. The code form
matches them to rounding, which shows the Table 3 values were fitted with the
code.

# Replicating Figure 2 (SM3 CalQD, regular spleen)

Figure 2 plots each compartment's concentration divided by the initial plasma
concentration (40 nM). The healthy panel runs to 4 h and the tumor panels to
12 h.

```{r figure2, fig.height = 4.5}
norm <- function(df) {
  df |>
    select(time, plasma, kidney, liver, spleen, lung, other, any_of(c("tumor", "breast"))) |>
    pivot_longer(-time, names_to = "compartment", values_to = "conc") |>
    mutate(rel = conc / 40)
}
fig2 <- bind_rows(lapply(names(stages), function(st) {
  name <- grid$model[grid$qd == "SM3 CalQD" & grid$stage == st]
  tmax <- if (st == "Healthy") 4 else 12
  norm(sims[[name]]) |>
    filter(time <= tmax, compartment != "plasma", compartment != "other") |>
    mutate(stage = factor(st, levels = names(stages)))
}))
ggplot(fig2, aes(time, rel, colour = compartment)) +
  geom_line() +
  facet_wrap(~stage, scales = "free_x") +
  labs(
    x = "Time (hours)", y = "Relative concentration (C / C_plasma,0)",
    colour = NULL,
    caption = "Replicates Figure 2 of Forder 2019."
  ) +
  theme_bw()
```

Figure 2 is described in the Results as follows. Spleen and kidney peak
quickly and then redistribute, the other tissues rise monotonically, the tumor
fills more slowly because its plasma flow is small, and in the early-stage
model the breast stays low relative to the tumor. The quantitative claim is
that the maximum SM3 CalQD tumor concentration is about 20% higher in the
late-stage model than in the early-stage model.

```{r figure2-claims}
tmax_of <- function(df, col) df$time[which.max(df[[col]])]
sm3 <- function(st) sims[[grid$model[grid$qd == "SM3 CalQD" & grid$stage == st]]]
claims <- tibble::tibble(
  stage = names(stages),
  spleen_tmax_h = vapply(names(stages), function(st) tmax_of(sm3(st), "spleen"), numeric(1)),
  kidney_tmax_h = vapply(names(stages), function(st) tmax_of(sm3(st), "kidney"), numeric(1)),
  tumor_cmax_nM = vapply(names(stages), function(st) {
    s <- sm3(st)
    if (is.null(s$tumor)) NA_real_ else max(s$tumor)
  }, numeric(1))
)
knitr::kable(
  claims |>
    rename(
      Stage = stage, "Spleen Tmax (h)" = spleen_tmax_h,
      "Kidney Tmax (h)" = kidney_tmax_h, "Tumor Cmax, 0-96 h (nM)" = tumor_cmax_nM
    ),
  digits = 3
)
tumor_ratio <- claims$tumor_cmax_nM[3] / claims$tumor_cmax_nM[2]
tumor_ratio
stopifnot(
  # Spleen and kidney peak within the first hour.
  all(claims$spleen_tmax_h < 1), all(claims$kidney_tmax_h < 1),
  # Paper: late-stage tumor Cmax ~20% higher than early-stage (1.23 here).
  tumor_ratio > 1.15, tumor_ratio < 1.3
)
```

The simulated late-to-early tumor Cmax ratio is 1.23, consistent with the
paper's "~20% higher". The early-stage value uses the Table 3 SM3 CalQD
partition coefficients discussed under Errata.

# Replicating Figure 3 (late-stage spleen, regular vs enlarged)

```{r figure3}
fig3 <- bind_rows(
  bind_rows(lapply(names(qds), function(q) {
    sims[[grid$model[grid$qd == q & grid$stage == "Late-stage"]]] |>
      mutate(qd = q, spleen_size = "Regular")
  })),
  enlarged |> filter(stage == "Late-stage") |> mutate(spleen_size = "Enlarged")
) |>
  filter(time <= 12)
ggplot(fig3, aes(time, spleen / 40, colour = qd, linetype = spleen_size)) +
  geom_line() +
  labs(
    x = "Time (hours)", y = "Relative spleen concentration",
    colour = NULL, linetype = "Spleen",
    caption = "Replicates Figure 3 of Forder 2019 (late-stage model)."
  ) +
  theme_bw()
```

As the paper notes, the early spleen peak disappears when the spleen volume is
doubled and its blood flow is kept the same.

# Replicating Figure S3 (SM3 CalQD over 4 days)

```{r figureS3, fig.height = 4.5}
figS3 <- bind_rows(lapply(names(stages), function(st) {
  norm(sm3(st)) |>
    filter(compartment != "plasma", compartment != "other") |>
    mutate(stage = factor(st, levels = names(stages)))
}))
ggplot(figS3, aes(time, rel, colour = compartment)) +
  geom_line() +
  facet_wrap(~stage) +
  labs(
    x = "Time (hours)", y = "Relative concentration (C / C_plasma,0)",
    colour = NULL, caption = "Replicates Figure S3 of Forder 2019."
  ) +
  theme_bw()
```

# PKNCA: tissue exposure over 0-96 h

The paper reports no NCA, so there is no published table to compare against.
PKNCA is used to summarise each tissue's Cmax, Tmax and AUC over 0-96 h for
the three SM3 CalQD models (the paper's main case). The grouping includes the
disease stage.

```{r pknca}
nca_in <- bind_rows(lapply(names(stages), function(st) {
  sm3(st) |>
    select(id, time, kidney, liver, spleen, lung, any_of("tumor")) |>
    pivot_longer(-c(id, time), names_to = "tissue", values_to = "conc") |>
    mutate(stage = st)
})) |>
  filter(!is.na(conc))

dose_df <- expand.grid(
  id = 1L, stage = names(stages),
  tissue = c("kidney", "liver", "spleen", "lung", "tumor"),
  stringsAsFactors = FALSE
) |>
  semi_join(distinct(nca_in, stage, tissue), by = c("stage", "tissue")) |>
  mutate(time = 0, amt = dose_pmol)

conc_obj <- PKNCA::PKNCAconc(nca_in, conc ~ time | stage + tissue / id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | stage + tissue + id, route = "intravascular")
intervals <- data.frame(start = 0, end = 96, cmax = TRUE, tmax = TRUE, auclast = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_tab <- as.data.frame(nca_res) |>
  select(stage, tissue, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
knitr::kable(
  nca_tab |>
    rename(
      Stage = stage, Tissue = tissue, "Cmax (nM)" = cmax,
      "Tmax (h)" = tmax, "AUC0-96 (nM*h)" = auclast
    ),
  digits = 3
)
stopifnot(nrow(nca_tab) == 14, !anyNA(nca_tab$cmax))
```

# Assumptions and deviations (Errata)

* **Deposited code is used where it differs from the printed equations.**
  The supplementary MATLAB code reproduces all nine Table 3 fits (and all six
  Table S2 fits) against Table 2. Where the code differs from the printed
  equations, the printed forms do not reproduce the fits, so the code is used:
  * *Late-stage plasma (Eq. 14).* The code returns
    `q_liver * liver / kp_liver + q_spleen * spleen / kp_spleen` to plasma. The
    printed Eq. 14 returns `(q_liver + q_spleen) * liver / kp_liver`. Spleen
    outflow is therefore counted twice in the code, since the liver equation
    receives it as well. This creates 1.0-3.6% spurious mass (depending on
    the QD) over the first hours (the net creation `q_spleen * (spleen/kp_spleen - liver/kp_liver)`
    goes to zero at equilibrium). The printed equation conserves mass but
    under-predicts Table 2 by about 3.5% for SM3 CalQD with the Table 3 partition coefficients
    (see "Mass balance").
  * *Early-stage breast and tumor (Eq. 19-20).* In the code the tumor is fed at
    the breast venous concentration `breast / kp_breast`, and the breast loses
    `q_tumor` at that same concentration. The printed equations drop
    `/ kp_breast` from both terms. Both forms conserve mass, but only the code
    form reproduces the early-stage ConQD and CalQD fits.
* **Early-stage SM3 CalQD partition coefficients.** The Table 3 column equals
  the Table S2 enlarged-spleen column, and these values reproduce Table 2
  only with a doubled spleen volume. The model ships them as printed, with
  the regular spleen volume, and they over-predict Table 2 by about 3%.
  Nothing was refitted. For the enlarged-spleen case, set
  `v_spleen = 0.154`.
* **Other-tissue flow.** The code uses `q_other` = 700.3091 mL/h for the
  healthy and late-stage models and 677.5191 mL/h for the early-stage model,
  where the 22.79 mL/h breast flow is subtracted. Table 1 prints only the
  early-stage 677.5, but its 944.3 mL/h plasma total equals the sum of flows
  with 700.3. The text says the breast *volume* is subtracted from `other`,
  but the code (and Table 1's 55.76 mL) keeps `v_other` unchanged in every
  model, and the code is what is followed here.
* **Liver elimination units.** Eq. 3 quotes `kelim` in 1/h, but Eq. 12 and
  the code multiply it by the liver concentration inside the `1/V_liver`
  bracket. It therefore behaves as a clearance of 1e-6 mL/h. Either way it is
  negligible over the simulated horizon.
* **Concentration scale.** The fit targets are fluorescence pixel intensities
  (Table 2). The code labels them nM and starts plasma at 40 nM, and that scale
  is kept here (dose 43.12 pmol into 1.078 mL). Because the model is linear,
  the dose only scales the output. The relative concentrations in Figures 2,
  3 and S3 do not depend on it.
* **Stray lines in the deposited code.** Both tumor-model scripts carry an
  active `vs = 2*0.077` (enlarged spleen) line, left over from the Table S2
  runs. The regular-spleen Table 3 fits reproduce only with `vs = 0.077`, so
  that value is used.
* **Sensitivity analysis not shipped as models.** The enlarged-spleen
  (Table S2) fits are reproduced above by `ini()` override, not packaged
  separately.
* **No variability.** The fits were relative least squares against mean
  intensities, with no uncertainty or variance components. The models are for
  typical-value simulation only.
