---
title: "Ceftazidime-avibactam in children (Franzese 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ceftazidime-avibactam in children (Franzese 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Franzese 2022 extended the adult ceftazidime-avibactam population PK models of
Li 2019 (see `vignette("Li_2019_ceftazidime_avibactam")` on the package site)
to children aged 3 months to < 18 years by adding PK data from one single-dose
phase I study and two multiple-dose phase II studies (complicated
intra-abdominal infection, cIAI, and complicated urinary tract infection,
cUTI). As in Li 2019, the two analytes were modelled separately -- two NONMEM
control streams, two data sets (9,628 observations from 2,130 subjects for
ceftazidime, 14,223 from 2,403 for avibactam) and two parameter tables -- so
the extraction is two model files sharing this vignette.

```{r models}
uiCaz <- rxode2::rxode(readModelDb("Franzese_2022_ceftazidime"))
uiAvi <- rxode2::rxode(readModelDb("Franzese_2022_avibactam"))
```

Both are two-compartment models with first-order elimination after IV
infusion. The pediatric adaptations shared by both analytes are body-weight
scaling of all four disposition parameters, BSA-normalized creatinine
clearance (NCrCL; bedside Schwartz in children, Cockcroft-Gault x 1.73 / BSA
in adults) for subjects older than 2 years, and a Rhodin renal-maturation
function of postmenstrual age that replaces NCrCL at 2 years and younger. The
ceftazidime weight effect on clearance is a normalized Emax function rather
than a power law.

Sources: the main text (Tables 1-4), the supplementary material
(CPT-111-635-s001: both final NONMEM control streams, the derivation of the
ceftazidime weight function, Tables S1-S3), and the 2024 erratum
(doi:10.1002/cpt.3143), which corrects the Table 2 column headings -- in the
original the ceftazidime and avibactam headings were transposed. No parameter
value is affected by the erratum.

# Population

```{r population, echo = FALSE}
tribble(
  ~Characteristic, ~`Phase I (any infection)`, ~`Phase II cIAI`, ~`Phase II cUTI`, ~Overall,
  "N", "32", "58", "63", "153",
  "Female, n (%)", "17 (53.1)", "16 (27.6)", "52 (82.5)", "85 (55.6)",
  "Age, years, median (range)", "5.7 (0.33-17.3)", "10.5 (3.00-17.0)", "3.8 (0.25-17.7)", "7.57 (0.25-17.7)",
  "Weight, kg, median (range)", "20.6 (5.4-60.5)", "37.9 (15.4-80.0)", "15.3 (4.1-71.0)", "25.0 (4.1-80.0)",
  "NCrCL, mL/min/1.73 m^2, median (range)", "130 (85.5-489)", "107 (59-271)", "89 (43-158)", "104 (43-489)",
  "Chinese, n (%)", "0", "6 (10.3)", "12 (19.0)", "18 (11.8)"
) |>
  knitr::kable(caption = "Pediatric baseline characteristics (Franzese 2022 Table 1).")
```

The 153 children were added to the adult data sets of Li 2019 (adults with
cIAI, cUTI or nosocomial pneumonia including ventilator-associated pneumonia,
subjects with renal impairment, and healthy volunteers). No child with
HAP/VAP was studied, and no child younger than 2 years was enrolled in the
cIAI trial. Pediatric doses (2-hour infusions, Table S1) were
50-12.5 mg/kg q8h (maximum 2,000-500 mg) for children 6 months and older and
40-10 mg/kg q8h for 3 to < 6 months, halved for CrCL 30 to < 50 mL/min.

# Source trace

```{r source-trace, echo = FALSE}
tribble(
  ~Quantity, ~Analyte, ~Value, ~Source,
  "CL, Vc, Q, Vp", "Ceftazidime", "7.75 L/h, 11.2 L, 5.33 L/h, 6.52 L", "Table S2 theta1-theta4",
  "NCrCL slopes on CL (hinge 100, cap 150)", "Ceftazidime", "0.0103036 / 0.00125182", "Table S2 'Slope 1' / 'Slope 2'; control stream SLOPE1, SLOPE2, NCLCR cap",
  "Emax weight effect on CL", "Ceftazidime", "WT50 = 53.5 kg, Hill 1, intercept 0", "Table S2 theta14, theta15; Supplementary Methods derivation; control stream TH13",
  "Weight exponents Q / Vc / Vp", "Ceftazidime", "0.67 / 1 / 1", "Results 'Ceftazidime'; control stream TH12, TH5",
  "cIAI and HAP/VAP on CL (bare multipliers)", "Ceftazidime", "1.33 / 1.10", "Table S2 theta16, theta17",
  "cUTI and cIAI-or-HAP/VAP on Vc (bare multipliers)", "Ceftazidime", "1.49 / 1.83", "Table S2 theta20, theta21",
  "Ventilator on Vc; Asian race on CL and Vc", "Ceftazidime", "+0.202; -0.136 (ASN), -0.0844 (CHN); -0.135 (Vc)", "Table S2 theta22, theta18, theta19, theta23",
  "IIV variances CL, Vc, Q, Vp", "Ceftazidime", "0.154, 0.108, 0.203, 0.0236", "Table S2 eta rows; control stream $OMEGA",
  "Residual SDs phase I / phase II-III", "Ceftazidime", "0.172 + 125 ng/mL / 0.374 + 2,560 ng/mL", "Table S2 theta8-theta11",
  "CL, Vc, Vp, Q", "Avibactam", "10.7 L/h, 11.5 L, 7.56 L, 6.94 L/h", "Table S3 theta1-theta4",
  "ESRD factor; dialysis CL", "Avibactam", "0.0674; 21.1 L/h", "Table S3 theta5, theta6",
  "NCrCL power below 80 / slope at or above 80 (cap 150)", "Avibactam", "0.986 / 0.00344", "Table S3 theta7, theta8; control stream CLCLCR",
  "Weight exponents CL, Q / Vc, Vp", "Avibactam", "0.67 / 1", "Results 'Avibactam'; control stream TH30, TH14",
  "Adult phase II cIAI on Vc and CL", "Avibactam", "+2.17 / +0.431", "Table S3 theta9, theta10",
  "cUTI on Vc; phase III cIAI, HAP/VAP, pediatric cIAI on Vc", "Avibactam", "+0.412; +0.214", "Table S3 theta11, theta12; control stream V1POP",
  "APACHE II on CL; ventilator on Vc", "Avibactam", "-0.192; +0.267", "Table S3 theta15, theta28",
  "IIV 4x4 block CL, Vc, Vp, Q", "Avibactam", "variances 0.3453, 1.139, 1.156, 5.487", "Table S3 eta rows",
  "Residual SDs phase I / II / III", "Avibactam", "0.174 + 43.8 ng/mL / 0.498 / 0.364", "Table S3 theta17-theta20",
  "Renal maturation (AGE <= 2 y)", "Both", "TM50 47.7 weeks, Hill 3.4", "Methods 'Model development'; control streams",
  "Unbound fractions", "Both", "0.85 ceftazidime, 0.92 avibactam", "Methods 'Simulations and PK/PD targets'"
) |>
  knitr::kable(caption = "Where every parameter and equation comes from.")
```

# Covariate-effect reproduction

Each model is solved for one typical subject (no IIV) at a reference
covariate setting and at a perturbed one; the ratio of the `cl` or `vc` the
model computes is compared with the value implied by the source. This has no
Monte Carlo component and is asserted tightly.

```{r covcheck}
baseCov <- data.frame(
  AGE = 30, PAGE = 30 * 52.18 + 40, WT = 70, CRCL = 80,
  DIS_CIAI = 0, DIS_CUTI = 0, DIS_HABP = 0, DIS_VABP = 0, MECH_VENT = 0,
  RACE_ASIAN_OTH = 0, RACE_CHINESE = 0, RACE_JAPANESE = 0,
  APACHE_II_SEV = 0, RENALIMP_ESRD = 0, RRT_HEMODIAL_STATUS = 0,
  STUDY_CIAI_PH2 = 0, STUDY_CAZAVI_PHASE2 = 0, STUDY_CAZAVI_PHASE3 = 1,
  STUDY_CAZAVI_PED_PHASE2 = 0
)

typicalParams <- function(ui, ...) {
  cov <- baseCov
  ch <- list(...)
  for (nm in names(ch)) cov[[nm]] <- ch[[nm]]
  cov <- cov[, intersect(names(cov), ui$all.covs), drop = FALSE]
  ev <- data.frame(
    id = 1L, time = c(0, 1), amt = c(1000, 0), evid = c(1L, 0L),
    dur = c(2, 0), cmt = "central"
  )
  ev <- cbind(ev, cov[rep(1L, 2L), , drop = FALSE])
  s <- rxode2::rxSolve(rxode2::zeroRe(ui), ev, returnType = "data.frame")
  c(cl = s$cl[1], vc = s$vc[1], q = s$q[1], vp = s$vp[1])
}

crefCaz <- 1 / 0.0103036 # CRCL where the ceftazidime spline equals 1
refCaz <- typicalParams(uiCaz, CRCL = crefCaz)
refAvi <- typicalParams(uiAvi)
emaxWt <- function(wt) wt / (53.5 + wt) * (53.5 + 70) / 70

checks <- tribble(
  ~Check, ~Analyte, ~Model, ~Expected,
  "CL at reference (L/h)", "Ceftazidime", unname(refCaz["cl"]), 7.75,
  "CL, CRCL 300 / CRCL 150 (cap)", "Ceftazidime",
  unname(typicalParams(uiCaz, CRCL = 300)["cl"] / typicalParams(uiCaz, CRCL = 150)["cl"]), 1,
  "CL, CRCL 150 / reference", "Ceftazidime",
  unname(typicalParams(uiCaz, CRCL = 150)["cl"] / refCaz["cl"]), 0.0103036 * 100 + 0.00125182 * 50,
  "CL, WT 53.5 / 70 kg (Emax at WT50)", "Ceftazidime",
  unname(typicalParams(uiCaz, CRCL = crefCaz, WT = 53.5)["cl"] / refCaz["cl"]), 0.5 * (1 + 53.5 / 70),
  "CL, WT 15 / 70 kg", "Ceftazidime",
  unname(typicalParams(uiCaz, CRCL = crefCaz, WT = 15)["cl"] / refCaz["cl"]), emaxWt(15),
  "CL, AGE 1 y at PMA 47.7 wk / reference", "Ceftazidime",
  unname(typicalParams(uiCaz, CRCL = crefCaz, AGE = 1, PAGE = 47.7)["cl"] / refCaz["cl"]), 0.5,
  "Q, WT 35 / 70 kg", "Ceftazidime",
  unname(typicalParams(uiCaz, CRCL = crefCaz, WT = 35)["q"] / refCaz["q"]), 0.5^0.67,
  "CL, cIAI", "Ceftazidime", unname(typicalParams(uiCaz, CRCL = crefCaz, DIS_CIAI = 1)["cl"] / refCaz["cl"]), 1.33,
  "CL, HAP", "Ceftazidime", unname(typicalParams(uiCaz, CRCL = crefCaz, DIS_HABP = 1)["cl"] / refCaz["cl"]), 1.1,
  "Vc, cUTI", "Ceftazidime", unname(typicalParams(uiCaz, CRCL = crefCaz, DIS_CUTI = 1)["vc"] / refCaz["vc"]), 1.49,
  "Vc, VAP + ventilator", "Ceftazidime",
  unname(typicalParams(uiCaz, CRCL = crefCaz, DIS_VABP = 1, MECH_VENT = 1)["vc"] / refCaz["vc"]), 1.83 * 1.202,
  "CL, Chinese", "Ceftazidime", unname(typicalParams(uiCaz, CRCL = crefCaz, RACE_CHINESE = 1)["cl"] / refCaz["cl"]), 1 - 0.0844,
  "CL, Japanese", "Ceftazidime", unname(typicalParams(uiCaz, CRCL = crefCaz, RACE_JAPANESE = 1)["cl"] / refCaz["cl"]), 1,
  "Vc, Japanese", "Ceftazidime", unname(typicalParams(uiCaz, CRCL = crefCaz, RACE_JAPANESE = 1)["vc"] / refCaz["vc"]), 1 - 0.135,
  "CL at reference (L/h)", "Avibactam", unname(refAvi["cl"]), 10.7,
  "CL, CRCL 150 / 80", "Avibactam", unname(typicalParams(uiAvi, CRCL = 150)["cl"] / refAvi["cl"]), 1 + 0.00344 * 70,
  "CL, CRCL 250 / 80 (cap)", "Avibactam", unname(typicalParams(uiAvi, CRCL = 250)["cl"] / refAvi["cl"]), 1 + 0.00344 * 70,
  "CL, CRCL 40 / 80", "Avibactam", unname(typicalParams(uiAvi, CRCL = 40)["cl"] / refAvi["cl"]), 0.5^0.986,
  "CL, ESRD (any CRCL)", "Avibactam", unname(typicalParams(uiAvi, CRCL = 5, RENALIMP_ESRD = 1)["cl"] / refAvi["cl"]), 0.0674,
  "CL, ESRD on hemodialysis (L/h)", "Avibactam",
  unname(typicalParams(uiAvi, CRCL = 5, RENALIMP_ESRD = 1, RRT_HEMODIAL_STATUS = 1)["cl"]), 21.1,
  "CL, WT 35 / 70 kg", "Avibactam", unname(typicalParams(uiAvi, WT = 35)["cl"] / refAvi["cl"]), 0.5^0.67,
  "Vp, WT 35 / 70 kg", "Avibactam", unname(typicalParams(uiAvi, WT = 35)["vp"] / refAvi["vp"]), 0.5,
  "CL, AGE 1 y at PMA 47.7 wk", "Avibactam",
  unname(typicalParams(uiAvi, AGE = 1, PAGE = 47.7, CRCL = 150)["cl"] / refAvi["cl"]), 0.5,
  "CL, APACHE II elevated", "Avibactam", unname(typicalParams(uiAvi, APACHE_II_SEV = 1)["cl"] / refAvi["cl"]), 1 - 0.192,
  "CL, adult phase II cIAI study", "Avibactam",
  unname(typicalParams(uiAvi, DIS_CIAI = 1, STUDY_CIAI_PH2 = 1)["cl"] / refAvi["cl"]), 1.431,
  "Vc, adult phase II cIAI study", "Avibactam",
  unname(typicalParams(uiAvi, DIS_CIAI = 1, STUDY_CIAI_PH2 = 1)["vc"] / refAvi["vc"]), 1 + 2.17,
  "Vc, phase III or pediatric cIAI", "Avibactam", unname(typicalParams(uiAvi, DIS_CIAI = 1)["vc"] / refAvi["vc"]), 1.214,
  "Vc, cUTI", "Avibactam", unname(typicalParams(uiAvi, DIS_CUTI = 1)["vc"] / refAvi["vc"]), 1.412,
  "Vc, HAP + ventilator", "Avibactam",
  unname(typicalParams(uiAvi, DIS_HABP = 1, MECH_VENT = 1)["vc"] / refAvi["vc"]), 1.214 * 1.267
) |>
  mutate(`Rel. diff` = Model / Expected - 1)

checks |>
  mutate(across(c(Model, Expected), \(x) signif(x, 5)), `Rel. diff` = signif(`Rel. diff`, 2)) |>
  knitr::kable(caption = "Typical-value covariate effects computed by the models against the source values.")

# Same parameters on both sides: pure numerical error, so a tight bound.
stopifnot(all(abs(checks$`Rel. diff`) < 1e-6))
```

# Virtual cohort

Franzese 2022 simulated 1,000 patients per indication, age group and dose
group, bootstrapping weight, age and NCrCL from a Pfizer database of 457
children with infections (older than 2 years) and taking weight from CDC
growth charts (2 years and younger), with between-subject variability
resampled from the post hoc random effects of the fitted subjects. None of
that subject-level material is published, so the cohort below is a
reconstruction (see Assumptions and deviations). Each arm has 100 subjects.

```{r cohort}
# set.seed() seeds R's RNG, not rxode2's; rxode2 streams are partitioned per
# solver thread, so a CI runner draws a different cohort than a workstation.
# Every assertion below is written to hold for any cohort the model produces.
set.seed(20220301)
rxode2::rxSetSeed(20220301)

N_ARM <- 100L

# Approximate median weight-for-age (kg), sexes pooled (growth-chart medians).
wfa <- data.frame(
  age = c(0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18),
  wt = c(6.0, 7.6, 8.9, 9.8, 11.0, 12.3, 14.3, 16.3, 18.4, 20.7, 23.0, 25.6, 28.6, 32.0, 36.0, 40.5, 45.5, 50.5, 55.0, 58.5, 61.0, 63.0)
)

ageGroups <- tribble(
  ~grp, ~lo, ~hi, ~mgkg,
  "12 to <18 years", 12, 18, 50,
  "6 to <12 years", 6, 12, 50,
  "2 to <6 years", 2, 6, 50,
  "1 to <2 years", 1, 2, 50,
  "6 to <12 months", 0.5, 1, 50,
  "3 to <6 months", 0.25, 0.5, 40,
  "Adults", 18, 80, NA
)

# Adult CrCL bands >= 80 mL/min from Li 2019 Table 3 (bottom block).
adultCrcl <- data.frame(lo = c(80, 150, 180), hi = c(150, 180, 610), n = c(955, 123, 116))

drawCrcl <- function(n, adult, renal) {
  if (renal == "Mild") {
    return(runif(n, 51, 80))
  }
  if (adult) {
    b <- sample(seq_len(nrow(adultCrcl)), n, replace = TRUE, prob = adultCrcl$n)
    return(runif(n, adultCrcl$lo[b], adultCrcl$hi[b]))
  }
  x <- rlnorm(n, log(125), 0.25)
  while (any(x < 80)) {
    low <- x < 80
    x[low] <- rlnorm(sum(low), log(125), 0.25)
  }
  x
}

makeArm <- function(grp, ind, renal, idOffset) {
  g <- ageGroups[ageGroups$grp == grp, ]
  adult <- grp == "Adults"
  age <- runif(N_ARM, g$lo, g$hi)
  wt <- if (adult) {
    pmin(pmax(rlnorm(N_ARM, log(70), 0.23), 40), 150)
  } else {
    approx(wfa$age, wfa$wt, age)$y * exp(rnorm(N_ARM, 0, ifelse(age < 2, 0.12, 0.18)))
  }
  race <- if (adult) {
    sample(c("other", "ASN", "CHN", "JPN"), N_ARM, replace = TRUE, prob = c(1209, 248, 262, 45))
  } else {
    sample(c("other", "ASN", "CHN"), N_ARM, replace = TRUE, prob = c(86.9, 1.3, 11.8))
  }
  vap <- if (ind == "HAP/VAP") rbinom(N_ARM, 1L, 138 / 413) else rep(0L, N_ARM)
  apache <- if (adult && ind != "cUTI") rbinom(N_ARM, 1L, 438 / 1115) else rep(0L, N_ARM)
  amtCaz <- if (adult) rep(2000, N_ARM) else pmin(g$mgkg * wt, 2000)
  data.frame(
    id = idOffset + seq_len(N_ARM), grp = grp, ind = ind, renal = renal,
    AGE = age, PAGE = age * 52.18 + 40, WT = wt, CRCL = drawCrcl(N_ARM, adult, renal),
    DIS_CIAI = as.integer(ind == "cIAI"), DIS_CUTI = as.integer(ind == "cUTI"),
    DIS_VABP = vap, DIS_HABP = as.integer(ind == "HAP/VAP") * (1L - vap),
    MECH_VENT = vap,
    RACE_ASIAN_OTH = as.integer(race == "ASN"), RACE_CHINESE = as.integer(race == "CHN"),
    RACE_JAPANESE = as.integer(race == "JPN"),
    APACHE_II_SEV = apache, RENALIMP_ESRD = 0L, RRT_HEMODIAL_STATUS = 0L,
    STUDY_CIAI_PH2 = 0L, STUDY_CAZAVI_PHASE2 = 0L,
    STUDY_CAZAVI_PHASE3 = as.integer(adult), STUDY_CAZAVI_PED_PHASE2 = as.integer(!adult),
    amtCaz = amtCaz, amtAvi = amtCaz / 4
  )
}

arms <- bind_rows(
  expand.grid(grp = ageGroups$grp, ind = c("cIAI", "cUTI", "HAP/VAP"), renal = "Normal", stringsAsFactors = FALSE),
  expand.grid(
    grp = c("12 to <18 years", "6 to <12 years", "2 to <6 years", "Adults"),
    ind = c("cIAI", "cUTI", "HAP/VAP"), renal = "Mild", stringsAsFactors = FALSE
  )
)
cohort <- bind_rows(lapply(seq_len(nrow(arms)), function(i) {
  makeArm(arms$grp[i], arms$ind[i], arms$renal[i], (i - 1L) * N_ARM)
}))

cohort |>
  group_by(Renal = renal, `Age group` = grp) |>
  summarise(
    N = n(), `Median WT (kg)` = round(median(WT), 1),
    `Median NCrCL` = round(median(CRCL)),
    `Median ceftazidime dose (mg)` = round(median(amtCaz)), .groups = "drop"
  ) |>
  knitr::kable(caption = "Virtual cohort by renal stratum and age group (all three indications pooled).")
```

# Simulation

Steady state is imposed with `ss = 1` on the first 2-hour infusion and two
further q8h doses are given explicitly, so the 0-24 h window is the paper's
`AUCss,0-24`. The first dosing interval is sampled every 0.1 h for the
time-above-threshold calculation; the rest of the window hourly. The paper's
PTA simulations used individual predictions, so residual error is switched
off (`sigma = NA`).

```{r simulate}
TINF <- 2
obsTimes <- sort(unique(c(seq(0, 8, by = 0.1), 9:24)))
covCols <- setdiff(names(cohort), c("grp", "ind", "renal", "amtCaz", "amtAvi"))

makeEvents <- function(amtCol) {
  dose <- cohort[rep(seq_len(nrow(cohort)), each = 3L), covCols]
  dose$time <- rep(c(0, 8, 16), nrow(cohort))
  dose$amt <- rep(cohort[[amtCol]], each = 3L)
  dose$evid <- 1L
  dose$dur <- TINF
  dose$ss <- rep(c(1L, 0L, 0L), nrow(cohort))
  dose$ii <- 8
  obs <- cohort[rep(seq_len(nrow(cohort)), each = length(obsTimes)), covCols]
  obs$time <- rep(obsTimes, nrow(cohort))
  obs$amt <- 0
  obs$evid <- 0L
  obs$dur <- 0
  obs$ss <- 0L
  obs$ii <- 0
  ev <- rbind(dose, obs)
  ev$cmt <- "central"
  ev[order(ev$id, ev$time, -ev$evid), ]
}

simOne <- function(ui, amtCol, drug) {
  s <- rxode2::rxSolve(ui, makeEvents(amtCol),
    sigma = NA, maxsteps = 500000L, returnType = "data.frame"
  )
  s |>
    transmute(id = as.integer(as.character(id)), time, Cc, drug = drug) |>
    left_join(cohort |> select(id, grp, ind, renal, WT), by = "id")
}

sim <- bind_rows(
  simOne(uiCaz, "amtCaz", "Ceftazidime"),
  simOne(uiAvi, "amtAvi", "Avibactam")
)

# The avibactam OMEGA block gives Q a variance of 5.5 (SD 2.3 on the log
# scale), so a handful of draws are extreme enough (Vc < 0.5 L with Q in the
# hundreds of L/h) that the steady-state solver gives up. Those subjects are
# dropped and counted; more than 1% would signal a real problem.
naIds <- unique(sim$id[sim$drug == "Avibactam" & !is.finite(sim$Cc)])
cat("Avibactam subjects with a failed steady-state solve:", length(naIds), "of", nrow(cohort), "\n")
stopifnot(
  all(is.finite(sim$Cc[sim$drug == "Ceftazidime"])),
  length(naIds) <= 0.01 * nrow(cohort)
)
sim <- sim |> filter(!(drug == "Avibactam" & id %in% naIds))
```

## Steady-state profiles

```{r profiles, fig.width = 7, fig.height = 5}
sim |>
  filter(renal == "Normal", grp %in% c("12 to <18 years", "2 to <6 years", "3 to <6 months", "Adults")) |>
  group_by(drug, ind, grp, time) |>
  summarise(med = median(Cc), lo = quantile(Cc, 0.05), hi = quantile(Cc, 0.95), .groups = "drop") |>
  ggplot(aes(time, med, colour = grp, fill = grp)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.1, colour = NA) +
  geom_line() +
  facet_grid(drug ~ ind, scales = "free_y") +
  scale_y_log10() +
  labs(
    x = "Time within the steady-state 24-hour window (h)",
    y = "Total plasma concentration (mg/L)", colour = "Age group", fill = "Age group"
  ) +
  theme_bw()
```

Median and 90% prediction interval at the approved doses, normal renal
function.

## Exposure by body weight

```{r figure1, fig.width = 7, fig.height = 5}
expo <- sim |>
  group_by(drug, id, grp, ind, renal, WT) |>
  summarise(cmax = max(Cc), .groups = "drop") |>
  filter(renal == "Normal")

expo |>
  mutate(population = ifelse(grp == "Adults", "Adults", "Children")) |>
  ggplot(aes(WT, cmax, colour = population)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE) +
  facet_grid(drug ~ ind, scales = "free_y") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Body weight (kg)", y = "Simulated Cmax,ss (mg/L)", colour = NULL) +
  theme_bw()
```

Replicates the simulated part of Figure 1b of Franzese 2022 (Cmax,ss against
body weight, children against adults with normal renal function): pediatric
peak concentrations sit at or above the adult band across the weight range,
the pattern the paper describes for its Figure 1 and Table 3.

# PKNCA validation

```{r pknca}
ncaConc <- sim |>
  filter(!is.na(Cc), time %in% c(0:24)) |>
  select(id, time, Cc, drug)
ncaDose <- sim |>
  filter(time == 0) |>
  distinct(id, drug) |>
  mutate(time = 0)

concObj <- PKNCA::PKNCAconc(ncaConc, Cc ~ time | drug + id, concu = "mg/L", timeu = "h")
doseObj <- PKNCA::PKNCAdose(ncaDose, ~ time | drug + id)
intervals <- data.frame(start = 0, end = 24, cmax = TRUE, auclast = TRUE)
ncaRes <- PKNCA::pk.nca(PKNCA::PKNCAdata(concObj, doseObj, intervals = intervals))

ncaInd <- as.data.frame(ncaRes$result) |>
  select(drug, id, PPTESTCD, PPORRES) |>
  left_join(cohort |> select(id, grp, ind, renal), by = "id")

stopifnot(all(is.finite(ncaInd$PPORRES)))

# Franzese 2022 tabulates geometric means.
ncaGm <- ncaInd |>
  group_by(drug, renal, ind, grp, PPTESTCD) |>
  summarise(PPORRES = exp(mean(log(PPORRES))), .groups = "drop")
```

`auclast` over the 0-24 h steady-state window is `AUCss,0-24`; `cmax` over
the window is `Cmax,ss` (end of infusion).

# Comparison against the published simulations (Table 3)

```{r table3}
grpLev <- c("12 to <18 years", "6 to <12 years", "2 to <6 years", "1 to <2 years", "6 to <12 months", "3 to <6 months", "Adults")
grpMild <- c("12 to <18 years", "6 to <12 years", "2 to <6 years", "Adults")
# Franzese 2022 Table 3, read column by column (cIAI, cUTI, HAP/VAP).
table3 <- bind_rows(
  tibble(drug = "Ceftazidime", renal = "Normal", grp = rep(grpLev, 3), ind = rep(c("cIAI", "cUTI", "HAP/VAP"), each = 7),
    cmax = c(64.6, 72.4, 68.2, 68.1, 72.1, 64.2, 58.9, 81.5, 91.5, 86.3, 86.0, 90.8, 80.7, 74.0, 71.8, 80.8, 76.4, 76.2, 80.4, 71.4, 65.1),
    auclast = c(618, 650, 572, 577, 637, 617, 602, 821, 864, 760, 767, 846, 820, 828, 747, 785, 691, 698, 769, 745, 712)),
  tibble(drug = "Avibactam", renal = "Normal", grp = rep(grpLev, 3), ind = rep(c("cIAI", "cUTI", "HAP/VAP"), each = 7),
    cmax = c(12.3, 14.2, 13.0, 13.6, 14.0, 12.1, 10.5, 11.9, 13.7, 12.5, 12.9, 13.3, 11.5, 9.73, 13.0, 15.1, 13.8, 14.4, 14.9, 12.9, 10.2),
    auclast = c(121, 136, 118, 125, 132, 121, 107, 121, 136, 118, 125, 132, 121, 113, 121, 136, 118, 125, 132, 121, 105)),
  tibble(drug = "Ceftazidime", renal = "Mild", grp = rep(grpMild, 3), ind = rep(c("cIAI", "cUTI", "HAP/VAP"), each = 4),
    cmax = c(81.6, 93.4, 88.2, 74.9, 103.0, 118.0, 111.0, 94.3, 90.9, 104.0, 98.0, 83.8),
    auclast = c(940, 1020, 892, 917, 1250, 1350, 1190, 1240, 1140, 1230, 1080, 1100)),
  tibble(drug = "Avibactam", renal = "Mild", grp = rep(grpMild, 3), ind = rep(c("cIAI", "cUTI", "HAP/VAP"), each = 4),
    cmax = c(14.7, 17.4, 15.9, 12.7, 14.2, 16.7, 15.3, 11.8, 15.5, 18.5, 17.0, 12.5),
    auclast = c(164, 192, 167, 148, 164, 192, 167, 152, 164, 192, 167, 147))
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = ncaGm, reference = table3,
  by = c("drug", "renal", "ind", "grp"),
  units = c(cmax = "mg/L", auclast = "mg*h/L, 0-24 h at steady state")
)
cmp |>
  dplyr::rename(Analyte = drug, `Renal function` = renal, Indication = ind, `Age group` = grp) |>
  knitr::kable(caption = "Geometric-mean steady-state exposures: Franzese 2022 Table 3 (Reference) against this simulation.")
```

```{r table3-assert}
pct <- ncaGm |>
  inner_join(
    table3 |> pivot_longer(c(cmax, auclast), names_to = "PPTESTCD", values_to = "ref"),
    by = c("drug", "renal", "ind", "grp", "PPTESTCD")
  ) |>
  mutate(pct_diff = 100 * (PPORRES / ref - 1))

pct |>
  group_by(Analyte = drug, Parameter = PPTESTCD) |>
  summarise(
    `Median % diff` = round(median(pct_diff), 1),
    `90th pct of |% diff|` = round(quantile(abs(pct_diff), 0.9), 1), .groups = "drop"
  ) |>
  knitr::kable(caption = "Agreement with Table 3 across the 33 arms.")

# The cohort is a reconstruction from marginal summaries, so the two sides
# differ by which subjects were drawn; assert on the centre and a robust
# quantile only. A mis-transcribed clearance, dose or unit moves the whole
# distribution by tens of percent.
caz <- pct[pct$drug == "Ceftazidime", ]
avi <- pct[pct$drug == "Avibactam", ]
stopifnot(
  abs(median(caz$pct_diff)) < 10,
  quantile(abs(caz$pct_diff), 0.9) < 20,
  # Avibactam runs high by a systematic, explained offset (next section).
  median(avi$pct_diff) > -10,
  median(avi$pct_diff) < 35
)
```

Ceftazidime reproduces Table 3 closely across all ages, indications and both
renal strata (median differences of a few percent on both `AUCss,0-24` and
`Cmax,ss`). Avibactam runs systematically high -- by roughly a tenth on
`AUCss,0-24` and a fifth on `Cmax,ss` -- and the explanation is in the next
section.

# Why the avibactam simulation runs high

Franzese 2022 did not simulate from the OMEGA matrix: it resampled each fitted
subject's post hoc random effects, pediatric ones for simulated children and
adult ones for simulated adults. If the fitted children's avibactam clearance
EBEs are centred above zero -- a pediatric offset the covariates do not
capture (the pediatric-study terms `CLPED` and `V1PED` were tested in the
control stream but held at zero in the final model) -- then the paper's
simulated children inherit it and a simulation centred on the typical value
does not.

Table 2 (as corrected by the erratum) makes the offset visible without any
knowledge of the cohort: in a child, ceftazidime and avibactam are dosed in a
4:1 ratio, so `(AUCcaz / AUCavi) / 4` is the subject's avibactam-to-ceftazidime
clearance ratio. The models' typical-value ratio for a child aged over 2
years with no infection-type effect on ceftazidime clearance can be computed
across the weights and NCrCL values that cover most of the children (10-45 kg;
NCrCL 90-150 mL/min/1.73 m^2, bracketing the Table 1 cohort medians of 89-130).
The ratio is somewhat higher right at the avibactam hinge of 80, where the
ceftazidime spline is lowest relative to its own reference.

```{r ratio}
ratioGrid <- expand.grid(WT = c(10, 15, 20, 30, 45), CRCL = c(90, 100, 125, 150))
ratioGrid$ratio <- mapply(function(wt, crcl) {
  typicalParams(uiAvi, WT = wt, CRCL = crcl, AGE = 8)["cl"] /
    typicalParams(uiCaz, WT = wt, CRCL = crcl, AGE = 8)["cl"]
}, ratioGrid$WT, ratioGrid$CRCL)

# Franzese 2022 Table 2 (erratum), phase I and phase II cUTI cohorts: no
# ceftazidime infection-type clearance effect applies to either.
table2 <- tribble(
  ~cohort, ~aucCaz, ~aucAvi,
  "Phase I, 3 months to <2 years", 888, 143,
  "Phase I, 2 to <6 years", 794, 127,
  "Phase I, 6 to <12 years", 814, 121,
  "Phase I, 12 to <18 years", 821, 124,
  "Phase II cUTI, 3 to <6 months", 736, 132,
  "Phase II cUTI, 6 months to <1 year", 859, 113,
  "Phase II cUTI, 1 to <2 years", 883, 117,
  "Phase II cUTI, 2 to <6 years", 789, 123,
  "Phase II cUTI, 6 to <12 years", 993, 153,
  "Phase II cUTI, 12 to <18 years", 843, 139
) |>
  mutate(ratio = aucCaz / aucAvi / 4)

tibble(
  Quantity = c(
    "Model typical CLavi / CLcaz, children > 2 y (range over WT 10-45 kg, NCrCL 90-150)",
    "Table 2 post hoc CLavi / CLcaz, median of 10 pediatric cohorts (range)"
  ),
  Value = c(
    sprintf("%.2f-%.2f", min(ratioGrid$ratio), max(ratioGrid$ratio)),
    sprintf("%.2f (%.2f-%.2f)", median(table2$ratio), min(table2$ratio), max(table2$ratio))
  )
) |>
  knitr::kable(caption = "Avibactam-to-ceftazidime clearance ratio: typical values against the fitted children.")

# Both sides are deterministic transcriptions / typical values.
stopifnot(median(table2$ratio) > max(ratioGrid$ratio))
```

The fitted children's median ratio exceeds the largest typical-value ratio
the two models produce over that covariate range. Because the ceftazidime simulation reproduces
Table 3 (which used the same resampling), the excess is on the avibactam side:
the fitted children's avibactam clearance is on average higher than the
avibactam typical value predicts, by roughly the amount the simulation above
overshoots Table 3 on `AUCss,0-24`. The larger overshoot on `Cmax,ss` points
to the same kind of offset in the resampled central-volume random effects. That is a property of the published model -- its
avibactam typical values are not centred on the pediatric data -- not an
extraction error, and a parametric simulation of children from this model
will over-predict avibactam exposure by about this much.

# Joint PK/PD target attainment (Table 4)

The joint target is 50% of the dosing interval with free ceftazidime above
the MIC (8 mg/L, `fu` = 0.85) and free avibactam above 1 mg/L (`fu` = 0.92),
achieved simultaneously.

```{r pta}
fT <- sim |>
  filter(time < 8) |>
  mutate(free = ifelse(drug == "Ceftazidime", 0.85 * Cc, 0.92 * Cc), thr = ifelse(drug == "Ceftazidime", 8, 1)) |>
  group_by(drug, id) |>
  summarise(ft = mean(free > thr), .groups = "drop") |>
  pivot_wider(names_from = drug, values_from = ft) |>
  filter(!is.na(Avibactam)) |>
  left_join(cohort |> select(id, grp, ind, renal), by = "id")

ptaSim <- fT |>
  group_by(renal, ind, grp) |>
  summarise(
    caz = 100 * mean(Ceftazidime >= 0.5), avi = 100 * mean(Avibactam >= 0.5),
    joint = 100 * mean(Ceftazidime >= 0.5 & Avibactam >= 0.5), .groups = "drop"
  )

table4 <- bind_rows(
  tibble(renal = "Normal", grp = rep(grpLev, 3), ind = rep(c("cIAI", "cUTI", "HAP/VAP"), each = 7),
    published = c(96, 90, 82, 82, 90, 93, 95, 99, 97, 94, 94, 98, 98, 97, 99, 97, 92, 92, 97, 98, 95)),
  tibble(renal = "Mild", grp = rep(grpMild, 3), ind = rep(c("cIAI", "cUTI", "HAP/VAP"), each = 4),
    published = c(99, 100, 100, 99, 99, 100, 100, 99, 99, 100, 100, 99))
)

ptaCmp <- inner_join(table4, ptaSim, by = c("renal", "ind", "grp"))
ptaCmp |>
  mutate(across(c(caz, avi, joint), round)) |>
  dplyr::rename(
    `Renal function` = renal, Indication = ind, `Age group` = grp,
    `Published joint PTA (%)` = published, `Ceftazidime PTA (%)` = caz,
    `Avibactam PTA (%)` = avi, `Joint PTA (%)` = joint
  ) |>
  knitr::kable(caption = "Joint PTA at MIC 8 mg/L: Franzese 2022 Table 4 against this simulation.")

stopifnot(
  # Centre: the parametric simulation sits below the published values (see
  # text), but by points, not tens of points.
  median(ptaCmp$joint - ptaCmp$published) > -15,
  median(ptaCmp$joint - ptaCmp$published) < 5,
  # Structural: the 33% higher ceftazidime CL in cIAI lowers attainment
  # against cUTI at the same ages. Pooled over the six pediatric age groups
  # (600 subjects per side) so the margin, about 9 points in development
  # runs, is several standard errors wide.
  mean(ptaCmp$joint[ptaCmp$renal == "Normal" & ptaCmp$ind == "cIAI" & ptaCmp$grp != "Adults"]) <
    mean(ptaCmp$joint[ptaCmp$renal == "Normal" & ptaCmp$ind == "cUTI" & ptaCmp$grp != "Adults"]),
  # Mild renal impairment raises exposure and therefore attainment.
  mean(ptaCmp$joint[ptaCmp$renal == "Mild"]) > mean(ptaCmp$joint[ptaCmp$renal == "Normal"])
)
```

Ceftazidime is the limiting analyte almost everywhere: avibactam attainment is
close to 100% in every arm. The simulation reproduces the paper's pattern --
attainment highest with mild renal impairment, lowest for cIAI children aged 1
to < 6 years, where the 33% higher ceftazidime clearance in cIAI bites
hardest -- but sits several points below Table 4. Two documented differences
push it that way: the paper resampled post hoc random effects in pairs from
the same fitted subject, which carries the positive correlation between the
two drugs' clearances into the joint target, whereas here the two models draw
independently; and the paper's resampled, shrinkage-reinflated random effects
are not the same distribution as the OMEGA matrix used here.

# Assumptions and deviations

1. **Ceftazidime clearance is the Table S2 value, 7.75 L/h.** The printed
   control stream's `$THETA` carries 9.1259 for CL while every other theta
   agrees with Table S2 to rounding. Table S2 is the final run (its footnote
   identifies it as run121, and the control stream's header says run121
   re-parameterized the weight function of run120, the likely source of the
   initial values). The simulated adult and pediatric ceftazidime AUCs above
   match Table 3 with 7.75; with 9.13 every arm would sit about 15% low.
2. **The avibactam control stream holds initial values.** Its `$THETA`
   matches the Li 2019 estimates and its `$OMEGA BLOCK(4)` is a 0.1 / 0.001
   starting guess; all avibactam values are from Table S3. Its functional
   forms are used as printed.
3. **Covariates held at zero in the final avibactam control stream** (sex,
   race, age powers, pediatric-study CL and Vc terms, the study-15 terms) are
   omitted from `model()`; sex, race and augmented renal clearance are listed
   under `covariatesDataExcluded`. The augmented-renal-clearance multiplier
   `TH13` is held at 1, so `RENAL_ARC` has no effect.
4. **Residual strata.** The avibactam `PH2` flag tests the adult phase 2 study
   numbers (`STDY = 2001/2002`), not `PHASE`, so pediatric phase II records
   fall in the avibactam phase I stratum. The ceftazidime stream tests
   `PHASE = 2 or 3`, which includes the pediatric phase II studies (the
   control streams' `STDY = 15` branch shows those studies carry
   `PHASE = 2`). The pediatric phase 2 studies therefore have their own
   indicator, `STUDY_CAZAVI_PED_PHASE2`, used only by the ceftazidime model.
5. **Pediatric phase I subjects** (suspected or confirmed infection) are
   taken to carry no infection-type indicator (`DIS_*` = 0); the paper does
   not state their `POP` code.
6. **The virtual cohort is a reconstruction.** Child weights are drawn around
   approximate growth-chart medians for age, child NCrCL (normal stratum)
   lognormally around 125 mL/min/1.73 m^2 truncated at 80 (the Table 1 cohort
   medians are 89-130), adult weights and NCrCL from the Li 2019 adult
   summaries, and race from Table 1 (children) and Li 2019 (adults). The
   Pfizer covariate database the paper bootstrapped is not public.
7. **Ventilation and APACHE II in the simulation.** As in the Li 2019
   vignette, a third of HAP/VAP subjects are ventilated (138 of 413 adult
   nosocomial-pneumonia subjects had VAP) and 39% of adult cIAI and HAP/VAP
   subjects have an elevated APACHE II score (438 of 1,115); children have no
   APACHE score.
8. **Between-subject variability** is drawn from the published OMEGA matrices,
   independently for the two analytes, whereas the paper resampled paired,
   shrinkage-reinflated post hoc random effects. This explains the avibactam
   exposure offset and the lower joint PTA discussed above. A few avibactam
   draws (extreme `Q` and `Vc`) fail the steady-state solve and are dropped.
9. **Table 2 headings.** The original Table 2 transposed the ceftazidime and
   avibactam column headings; the values used here follow the 2024 erratum.
