---
title: "Tacrolimus after kidney transplantation (Francke 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tacrolimus after kidney transplantation (Francke 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 96
)
```

```{r libraries, message = FALSE, warning = FALSE}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(ggplot2)
```

Francke and colleagues developed an international, multicentre population
pharmacokinetic model for twice-daily oral immediate-release tacrolimus in
adult kidney transplant recipients, together with a closed-form starting-dose
algorithm. Two models are distributed from this paper:

* `Francke_2025_tacrolimus` -- the **full model**, which uses the two
  post-transplant laboratory covariates (hematocrit, serum creatinine) in
  addition to genotype, age, and height. This is the follow-up dosing model.
* `Francke_2025_tacrolimus_startingdose` -- the **starting-dose model**, refit
  on the same data with only the covariates known *before* transplantation
  (genotype, age, height). This is the model behind the published dosing
  algorithm.

```{r load-models}
mod_full  <- rxode2::rxode(readModelDb("Francke_2025_tacrolimus"))
mod_start <- rxode2::rxode(readModelDb("Francke_2025_tacrolimus_startingdose"))

mod_full_typ  <- rxode2::zeroRe(mod_full)
mod_start_typ <- rxode2::zeroRe(mod_start)
```

# Population

```{r population}
pop <- mod_full$population
tibble::tibble(
  Field = names(pop),
  Value = vapply(pop, function(x) paste(as.character(x), collapse = "; "), character(1))
) |>
  knitr::kable()
```

The analysis pooled 13,427 whole-blood tacrolimus concentrations from 1180
adult kidney transplant recipients at four European centres (Erasmus MC
Rotterdam n = 547, LUMC Leiden n = 100, Bellvitge Barcelona n = 444, St. Luc
Brussels n = 89). Of these concentrations 11,670 were pre-dose (trough)
samples; the remaining 1757 came from 208 densely sampled AUC profiles in 185
recipients. Baseline characteristics are Francke 2025 Table 1.

# Source trace

Every structural parameter, covariate coefficient, and variance term in both
model files, with the location in the source it was taken from. Francke 2025
publishes a complete NONMEM control stream as Supplementary Data S3, which is
cited alongside Table 2 wherever it independently confirms a value or resolves
a scale ambiguity.

```{r source-trace}
source_trace <- tibble::tribble(
  ~Component,                    ~`Full model`, ~`Starting-dose model`, ~Source,
  "Structural model",            "2-cmt, 1st-order abs + lag", "2-cmt, 1st-order abs + lag", "Section 3.2.1; Suppl. S3 $SUBROUTINE ADVAN4 TRANS4",
  "tlag (h)",                    "0.375",  "0.375",  "Table 2; Suppl. S3 THETA(1)",
  "ka (1/h)",                    "6.59",   "6.25",   "Table 2; Suppl. S3 THETA(2). Table header 'K a (L/h)' is a typo; Abstract says 6.59/h",
  "CL/F (L/h)",                  "20.7",   "19.9",   "Table 2; Suppl. S3 THETA(4)",
  "V1/F (L)",                    "705",    "653",    "Table 2; Suppl. S3 THETA(3)",
  "Q/F (L/h)",                   "8.54",   "9.14",   "Table 2; Suppl. S3 THETA(6)",
  "V2/F (L), FIXED",             "7670",   "7670",   "Table 2 '(f)'; Section 3.2.1; Suppl. S3 THETA(5) = 7670 FIX",
  "CYP3A5 *1/*3 on CL/F",        "1.64",   "1.64",   "Table 2; Section 3.2.2 equation; Suppl. S3 THETA(12)",
  "CYP3A5 *1/*1 on CL/F",        "1.93",   "1.9",    "Table 2; Section 3.2.2 equation; Suppl. S3 THETA(13)",
  "CYP3A4*22 on CL/F",           "0.836",  "0.842",  "Table 2; Section 3.2.2 equation; Suppl. S3 THETA(18)",
  "Hematocrit exponent on CL/F", "-0.51 (centred 0.33 L/L)", "not in model", "Table 2; Suppl. S3 THETA(14), CLHCT = (HCT/0.33)**THETA(14)",
  "Age exponent on CL/F",        "-0.309 (centred 57.5 y)",  "-0.332 (centred 57.5 y)", "Table 2; Suppl. S3 THETA(15), CLAGE = (AGE/57.5)**THETA(15)",
  "Creatinine exponent on CL/F", "-0.0905 (centred 147 umol/L)", "not in model", "Table 2; Suppl. S3 THETA(16), CLCREAT = (CREAT/147)**THETA(16)",
  "Height exponent on CL/F",     "1.17 (centred 170 cm)",    "0.97 (centred 170 cm)", "Table 2; Suppl. S3 THETA(17), CLHGT = (HGT/170)**THETA(17)",
  "IIV CL/F",                    "41.4% CV",  "41.8% CV",  "Table 2 'IIV (%)'; Suppl. S3 $OMEGA BLOCK(3) ETA(1)",
  "IIV V1/F",                    "80.5% CV",  "77.8% CV",  "Table 2 'IIV (%)'; Suppl. S3 $OMEGA BLOCK(3) ETA(2)",
  "IIV Q/F",                     "82.8% CV",  "81% CV",    "Table 2 'IIV (%)'; Suppl. S3 $OMEGA BLOCK(3) ETA(3)",
  "Correlations CL-V1 / CL-Q / V1-Q", "58% / 6.5% / 12.5%", "58.4% / 7.9% / 14%", "Table 2 'Correlation matrix'",
  "Residual, Rotterdam immunoassay",  "0.242", "0.245", "Table 2; Suppl. S3 THETA(7)",
  "Residual, Rotterdam+Leiden LC-MS/MS", "0.281", "0.284", "Table 2; Suppl. S3 THETA(8)",
  "Residual, Barcelona immunoassay",   "0.376", "0.381", "Table 2; Suppl. S3 THETA(9)",
  "Residual, Barcelona LC-MS/MS",      "0.419", "0.419", "Table 2; Suppl. S3 THETA(10)",
  "Residual, Brussels immunoassay",    "0.313", "0.323", "Table 2; Suppl. S3 THETA(11)"
)
knitr::kable(source_trace)
```

## Two scale ambiguities the supplement resolves

**The IIV column is %CV, not a variance.** Table 2 reports base-model IIV on
CL/F as 51.3% and final-model IIV as 41.4%, without stating the scale. Reading
the column as %CV reproduces *both* summary statistics the paper quotes
elsewhere, which a variance reading cannot:

```{r iiv-scale-check}
pct_cv  <- function(base, final) 100 * (base - final) / base
pct_var <- function(base, final) 100 * (base^2 - final^2) / base^2

scale_check <- tibble::tibble(
  Statistic = c(
    "Full model, reduction on the CV scale",
    "Full model, reduction on the variance scale",
    "Starting-dose model, reduction on the variance scale",
    "Andrews 2017 full model, reduction on the variance scale"
  ),
  `Table 2 inputs` = c(
    "51.3% -> 41.4%", "51.3% -> 41.4%", "51.3% -> 41.8%", "46.3% -> 38.6%"
  ),
  Computed = sprintf("%.1f%%", c(
    pct_cv(51.3, 41.4),
    pct_var(51.3, 41.4),
    pct_var(51.3, 41.8),
    pct_var(46.3, 38.6)
  )),
  `Published in Francke 2025` = c(
    "19.3% (Abstract)",
    "35% (Key Points, Section 3.2.2, Conclusions)",
    "33.5% (Discussion paragraph 6)",
    "30.4% (Discussion paragraph 6)"
  )
)
knitr::kable(scale_check)

# Every published percentage is recovered under the %CV reading.
stopifnot(abs(pct_cv(51.3, 41.4)  - 19.3) < 0.1)
stopifnot(abs(pct_var(51.3, 41.4) - 35.0) < 0.2)
stopifnot(abs(pct_var(51.3, 41.8) - 33.5) < 0.2)
stopifnot(abs(pct_var(46.3, 38.6) - 30.4) < 0.2)
```

Four separate published percentages -- spanning both of this paper's models
*and* the Andrews 2017 comparator quoted in the Discussion -- are all
recovered from the Table 2 column under the %CV reading. A variance reading
recovers none of them. The model files therefore convert to log-scale
variances with `omega^2 = log(1 + CV^2)`, the same convention used by
`Andrews_2017_tacrolimus` from the same research group.

**The residual column holds standard deviations on the natural-log scale.**
Section 2.2.1 says only that "a logarithmic error model" was used. The
supplement is explicit: `$SIGMA` is `1 FIX` for all five error terms, and
`$ERROR` computes `IPRED = LOG(F)` followed by `Y = IPRED + ERR1*EPS(1)` with
`ERR1 = SQRT(THETA(7)**2)`. With EPS fixed at unit variance, each THETA *is*
the residual SD on the log scale, which maps onto nlmixr2's `lnorm()` family.

# Virtual cohort

Covariate distributions follow Francke 2025 Table 1 (Total column). Table 1
reports medians and interquartile ranges, so continuous covariates are drawn
from log-normal distributions matched to the reported median and IQR.

```{r cohort-helpers}
# Match a lognormal to a reported median and IQR.
rlnorm_iqr <- function(n, med, q25, q75) {
  sdlog <- (log(q75) - log(q25)) / (2 * stats::qnorm(0.75))
  stats::rlnorm(n, meanlog = log(med), sdlog = sdlog)
}

# The full model needs all seven covariates; the starting-dose model needs a
# subset. Supplying the full set to either is harmless.
COVS <- c("AGE", "HT", "HCT", "CREAT",
          "CYP3A5_STAR1_HET", "CYP3A5_STAR1_HOM",
          "SNP_CYP3A4_RS35599367", "STUDY_TACRO_FRANCKE")

# Build an event table for twice-daily dosing evaluated at steady state.
# ss = 1 gives the analytic steady state, which matters here because the
# model's terminal half-life is very long (see below).
make_ss_events <- function(subj, dose_col = "dose", tau = 12,
                           obs_times = seq(0, 12, by = 0.1)) {
  stopifnot("id" %in% names(subj))
  dosing <- subj |>
    dplyr::mutate(time = 0, amt = .data[[dose_col]], evid = 1L,
                  cmt = "depot", ss = 1L, ii = tau)
  # Observations sit on the ODE state `central`; rxode2 returns the algebraic
  # observable Cc as a column at those rows.
  obs <- subj |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(amt = NA_real_, evid = 0L,
                  cmt = "central", ss = 0L, ii = 0)
  dplyr::bind_rows(dosing, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    as.data.frame()
}

# The default liblsoda integrator fails on this model's ss = 1 records
# (the fixed 7670 L peripheral volume makes the steady state very stiff);
# lsoda solves it cleanly.
solve_ss <- function(mod, events, keep = character()) {
  out <- rxode2::rxSolve(mod, events = events, method = "lsoda",
                         keep = keep) |>
    as.data.frame()
  # rxSolve omits `id` entirely for a single-subject event table.
  if (is.null(out$id)) out$id <- 1L
  out
}
```

## The model has a very long terminal half-life

`V2/F` is fixed at 7670 L against a `Q/F` of only 8.54 L/h, so the deep
peripheral compartment turns over slowly. This matters for every steady-state
statement below, and it is why the simulations use analytic steady state
rather than a finite run-in.

```{r half-life}
cl <- 20.7; vc <- 705; q <- 8.54; vp <- 7670
kel <- cl / vc; k12 <- q / vc; k21 <- q / vp
a <- kel + k12 + k21
b <- kel * k21
alpha <- (a + sqrt(a^2 - 4 * b)) / 2
beta  <- (a - sqrt(a^2 - 4 * b)) / 2

tibble::tibble(
  Phase = c("Distribution (alpha)", "Terminal (beta)"),
  `Half-life (h)`    = round(log(2) / c(alpha, beta), 1),
  `Half-life (days)` = round(log(2) / c(alpha, beta) / 24, 2)
) |>
  knitr::kable()

sprintf("Vss/F = %.0f L; ~5 terminal half-lives = %.0f days to steady state.",
        vc + vp, 5 * log(2) / beta / 24)
```

# Replicate published figures

Francke 2025 Figure 3 shows simulated steady-state whole-blood concentrations
as each final-model covariate is varied while the others are held at the
population median. Table 1 reports medians and interquartile ranges but not
the 10th and 90th percentiles, so the panels below use the 25th percentile,
median, and 75th percentile of each covariate.

```{r figure-3-reference}
# Reference subject: every covariate multiplier on CL/F equals exactly 1.
REF <- list(AGE = 57.5, HT = 170, HCT = 0.33, CREAT = 147,
            CYP3A5_STAR1_HET = 0, CYP3A5_STAR1_HOM = 0,
            SNP_CYP3A4_RS35599367 = 0, STUDY_TACRO_FRANCKE = 3)

# A 5 mg twice-daily dose, close to the cohort median of 4 mg (Table 1),
# is held constant across every panel so the panels show the covariate
# effect on exposure rather than a dose effect.
PANEL_DOSE <- 5

make_panel_subjects <- function(varying_name, values, labels) {
  dplyr::bind_rows(lapply(seq_along(values), function(i) {
    row <- REF
    row[[varying_name]] <- values[[i]]
    tibble::as_tibble(row) |>
      dplyr::mutate(id = i, level = labels[i], dose = PANEL_DOSE)
  }))
}

panel_plot <- function(varying_name, values, labels, title, legend_title) {
  subj <- make_panel_subjects(varying_name, values, labels)
  sim  <- solve_ss(mod_full_typ, make_ss_events(subj), keep = "level")
  sim |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::mutate(level = factor(level, levels = labels)) |>
    ggplot(aes(time, Cc, colour = level)) +
    geom_line(linewidth = 0.9) +
    labs(x = "Time after dose at steady state (h)",
         y = "Tacrolimus whole blood (ng/mL)",
         colour = legend_title, title = title,
         caption = "Replicates Francke 2025 Figure 3.") +
    expand_limits(y = 0) +
    theme_minimal()
}
```

```{r figure-3a, fig.cap = "Figure 3A: CYP3A5 genotype."}
subj_3a <- dplyr::bind_rows(
  tibble::as_tibble(REF) |> dplyr::mutate(id = 1, level = "CYP3A5 *3/*3"),
  tibble::as_tibble(REF) |> dplyr::mutate(id = 2, level = "CYP3A5 *1/*3",
                                          CYP3A5_STAR1_HET = 1),
  tibble::as_tibble(REF) |> dplyr::mutate(id = 3, level = "CYP3A5 *1/*1",
                                          CYP3A5_STAR1_HOM = 1)
) |>
  dplyr::mutate(dose = PANEL_DOSE)

sim_3a <- solve_ss(mod_full_typ, make_ss_events(subj_3a), keep = "level")
sim_3a |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(level = factor(level, levels = c("CYP3A5 *3/*3", "CYP3A5 *1/*3", "CYP3A5 *1/*1"))) |>
  ggplot(aes(time, Cc, colour = level)) +
  geom_line(linewidth = 0.9) +
  labs(x = "Time after dose at steady state (h)",
       y = "Tacrolimus whole blood (ng/mL)", colour = "CYP3A5 genotype",
       title = "Figure 3A: CYP3A5 genotype effect",
       caption = "Replicates Francke 2025 Figure 3A.") +
  expand_limits(y = 0) +
  theme_minimal()
```

```{r figure-3b, fig.cap = "Figure 3B: hematocrit."}
panel_plot("HCT", c(0.300, 0.332, 0.370),
           c("0.300 L/L (25th)", "0.332 L/L (median)", "0.370 L/L (75th)"),
           "Figure 3B: hematocrit effect", "Hematocrit")
```

```{r figure-3c, fig.cap = "Figure 3C: age."}
panel_plot("AGE", c(45, 57, 65),
           c("45 y (25th)", "57 y (median)", "65 y (75th)"),
           "Figure 3C: age effect", "Age")
```

```{r figure-3d, fig.cap = "Figure 3D: serum creatinine."}
panel_plot("CREAT", c(110, 140, 191),
           c("110 umol/L (25th)", "140 umol/L (median)", "191 umol/L (75th)"),
           "Figure 3D: serum creatinine effect", "Serum creatinine")
```

```{r figure-3e, fig.cap = "Figure 3E: height."}
panel_plot("HT", c(162, 170, 177),
           c("162 cm (25th)", "170 cm (median)", "177 cm (75th)"),
           "Figure 3E: height effect", "Height")
```

```{r figure-3f, fig.cap = "Figure 3F: CYP3A4 genotype."}
subj_3f <- dplyr::bind_rows(
  tibble::as_tibble(REF) |> dplyr::mutate(id = 1, level = "CYP3A4*1/*1"),
  tibble::as_tibble(REF) |> dplyr::mutate(id = 2, level = "CYP3A4*22 carrier",
                                          SNP_CYP3A4_RS35599367 = 1)
) |>
  dplyr::mutate(dose = PANEL_DOSE)

sim_3f <- solve_ss(mod_full_typ, make_ss_events(subj_3f), keep = "level")
sim_3f |>
  dplyr::filter(!is.na(Cc)) |>
  ggplot(aes(time, Cc, colour = level)) +
  geom_line(linewidth = 0.9) +
  labs(x = "Time after dose at steady state (h)",
       y = "Tacrolimus whole blood (ng/mL)", colour = "CYP3A4 genotype",
       title = "Figure 3F: CYP3A4*22 effect",
       caption = "Replicates Francke 2025 Figure 3F.") +
  expand_limits(y = 0) +
  theme_minimal()
```

Each panel reproduces the direction the paper reports (Section 3.3): higher
age, hematocrit, and serum creatinine raise exposure at a fixed dose (i.e.
lower the dose requirement), taller recipients need more drug, CYP3A5
expressers need more, and CYP3A4*22 carriers need less.

# PKNCA validation

Steady-state non-compartmental analysis over one 12-hour dosing interval, by
CYP3A5 genotype. A time-zero record is present by construction, so no
defensive row is needed.

```{r pknca}
subj_nca <- subj_3a
sim_nca  <- solve_ss(mod_full_typ, make_ss_events(subj_nca, obs_times = seq(0, 12, by = 0.05)),
                     keep = "level")

conc_df <- sim_nca |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, genotype = level)

dose_df <- subj_nca |>
  dplyr::transmute(id, time = 0, amt = dose, genotype = level)

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | genotype + id,
                             concu = "ng/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | genotype + id,
                             doseu = "mg")

intervals <- data.frame(
  start = 0, end = 12,
  cmax = TRUE, tmax = TRUE, cmin = TRUE,
  auclast = TRUE, cav = TRUE
)

res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_tbl <- as.data.frame(res$result)

nca_wide <- nca_tbl |>
  dplyr::select(genotype, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

nca_wide |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 2))) |>
  dplyr::rename(
    "CYP3A5 genotype" = genotype,
    "Cmax (ng/mL)"    = cmax,
    "Tmax (h)"        = tmax,
    "Cmin (ng/mL)"    = cmin,
    "AUC0-12 (ng*h/mL)" = auclast,
    "Cavg (ng/mL)"    = cav
  ) |>
  knitr::kable(caption = "Steady-state NCA at 5 mg twice daily, by CYP3A5 genotype.")
```

# Comparison against published values

## Structural identity: AUC0-tau at steady state equals Dose / (CL/F)

This is the identity the published dosing algorithm inverts, so it must hold
per subject before the algorithm can be checked.

```{r auc-identity}
cl_of <- function(cyp3a5_het, cyp3a5_hom, cyp3a4s22, hct, age, creat, ht) {
  20.7 *
    (1 + (1.64 - 1) * cyp3a5_het + (1.93 - 1) * cyp3a5_hom) *
    (1 + (0.836 - 1) * cyp3a4s22) *
    (hct / 0.33)^(-0.51) *
    (age / 57.5)^(-0.309) *
    (creat / 147)^(-0.0905) *
    (ht / 170)^(1.17)
}

identity_tbl <- nca_wide |>
  dplyr::mutate(
    het = as.integer(genotype == "CYP3A5 *1/*3"),
    hom = as.integer(genotype == "CYP3A5 *1/*1"),
    `CL/F (L/h)` = cl_of(het, hom, 0, 0.33, 57.5, 147, 170),
    `Dose / (CL/F) (ng*h/mL)` = PANEL_DOSE / `CL/F (L/h)` * 1000,
    `Ratio (NCA / identity)`  = auclast / `Dose / (CL/F) (ng*h/mL)`
  )

identity_tbl |>
  dplyr::transmute(
    "CYP3A5 genotype" = genotype,
    "CL/F (L/h)" = round(`CL/F (L/h)`, 2),
    "NCA AUC0-12 (ng*h/mL)" = round(auclast, 1),
    "Dose / (CL/F) (ng*h/mL)" = round(`Dose / (CL/F) (ng*h/mL)`, 1),
    "Ratio" = round(`Ratio (NCA / identity)`, 4)
  ) |>
  knitr::kable()

stopifnot(nrow(identity_tbl) == 3L)
stopifnot(all(abs(identity_tbl$`Ratio (NCA / identity)` - 1) < 0.005))
```

The identity holds to better than 0.5% for all three genotypes, which
validates the clearance covariate model, the `S2 = V2/1000` dose scaling that
puts `Cc` in ng/mL, and the reference covariate set simultaneously.

## Published dose-requirement ratios

Francke 2025 Discussion paragraph 3 and the Key Points state that CYP3A5
*1/*3 recipients need roughly 1.7 times, and *1/*1 recipients roughly 2 times,
the dose of *3/*3 non-expressers, while CYP3A4*22 carriers need about 0.8
times the dose of CYP3A4*1/*1 recipients. Because dose requirement is
proportional to CL/F, these are read directly off the fitted multipliers.

```{r dose-ratios}
ratio_tbl <- tibble::tibble(
  Contrast = c("CYP3A5 *1/*3 vs *3/*3", "CYP3A5 *1/*1 vs *3/*3",
               "CYP3A4*22 carrier vs *1/*1"),
  `Model CL/F ratio` = c(
    cl_of(1, 0, 0, 0.33, 57.5, 147, 170) / cl_of(0, 0, 0, 0.33, 57.5, 147, 170),
    cl_of(0, 1, 0, 0.33, 57.5, 147, 170) / cl_of(0, 0, 0, 0.33, 57.5, 147, 170),
    cl_of(0, 0, 1, 0.33, 57.5, 147, 170) / cl_of(0, 0, 0, 0.33, 57.5, 147, 170)
  ),
  `Published statement` = c("1.7 times higher", "2 times higher", "0.8 times lower")
)

ratio_tbl |>
  dplyr::mutate(`Model CL/F ratio` = round(`Model CL/F ratio`, 3)) |>
  knitr::kable()

stopifnot(abs(ratio_tbl$`Model CL/F ratio`[1] - 1.64)  < 1e-6)
stopifnot(abs(ratio_tbl$`Model CL/F ratio`[2] - 1.93)  < 1e-6)
stopifnot(abs(ratio_tbl$`Model CL/F ratio`[3] - 0.836) < 1e-6)
```

## The published starting-dose algorithm

Section 3.4 gives the algorithm in closed form. Reproduced exactly as printed:

$$
\text{Dose (mg)} = \frac{234 \times 20.7 \times f_{CYP3A5} \times f_{CYP3A4}
  \times \left(\frac{\text{Age}}{57.5}\right)^{-0.332}
  \times \left(\frac{\text{Height}}{170}\right)^{0.97}}{1000}
$$

with `f_CYP3A5` = 1.0 / 1.64 / 1.9 for *3/*3 / *1/*3 / *1/*1 and `f_CYP3A4` =
1.0 / 0.842 for *1 / *22. The constant 234 ng*h/mL is the AUC0-12h the paper
states corresponds to a 10 ng/mL pre-dose target.

```{r dose-algorithm}
# Reproduced verbatim from Section 3.4, including the 20.7 constant. Note that
# 20.7 is the FULL model's CL/F; the starting-dose model's own CL/F is 19.9.
francke_starting_dose <- function(age, height,
                                  cyp3a5 = c("*3/*3", "*1/*3", "*1/*1"),
                                  cyp3a4_22 = FALSE,
                                  auc_target = 234, cl_constant = 20.7) {
  cyp3a5 <- match.arg(cyp3a5)
  f_cyp3a5 <- c("*3/*3" = 1.0, "*1/*3" = 1.64, "*1/*1" = 1.9)[[cyp3a5]]
  f_cyp3a4 <- if (isTRUE(cyp3a4_22)) 0.842 else 1.0
  auc_target * cl_constant * f_cyp3a5 * f_cyp3a4 *
    (age / 57.5)^(-0.332) * (height / 170)^(0.97) / 1000
}

dose_examples <- tidyr::expand_grid(
  age = c(45, 57.5, 65),
  height = c(162, 170, 177),
  cyp3a5 = c("*3/*3", "*1/*3", "*1/*1")
) |>
  dplyr::mutate(`Dose (mg twice daily)` = round(mapply(
    francke_starting_dose, age, height, cyp3a5), 2))

dose_examples |>
  dplyr::rename("Age (y)" = age, "Height (cm)" = height,
                "CYP3A5 genotype" = cyp3a5) |>
  knitr::kable(caption = "Starting doses from the published algorithm.")

sprintf("Reference recipient (57.5 y, 170 cm, CYP3A5 *3/*3, CYP3A4*1/*1): %.3f mg twice daily.",
        francke_starting_dose(57.5, 170))
```

## Does the algorithm dose actually produce AUC0-12 = 234 ng*h/mL?

```{r algorithm-auc-check}
alg_dose <- francke_starting_dose(57.5, 170)

subj_alg <- tibble::as_tibble(REF) |>
  dplyr::mutate(id = 1L, dose = alg_dose, level = "algorithm dose")

sim_alg <- solve_ss(mod_full_typ,
                    make_ss_events(subj_alg, obs_times = seq(0, 12, by = 0.02)),
                    keep = "level") |>
  dplyr::filter(!is.na(Cc))

auc_alg  <- with(sim_alg, sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2))
ctrough  <- sim_alg$Cc[nrow(sim_alg)]
c0       <- sim_alg$Cc[1]

check_tbl <- tibble::tibble(
  Quantity = c("Dose (mg twice daily)", "AUC0-12 at steady state (ng*h/mL)",
               "Pre-dose concentration (ng/mL)", "AUC0-12 / pre-dose ratio"),
  `Simulated` = c(round(alg_dose, 3), round(auc_alg, 1),
                  round(c0, 2), round(auc_alg / c0, 1)),
  `Francke 2025` = c("4.844 (from the printed formula)", "234", "10", "23.4")
)
knitr::kable(check_tbl)

# The AUC target is reproduced essentially exactly.
stopifnot(abs(auc_alg / 234 - 1) < 0.005)
```

The AUC leg of the algorithm reproduces to within 0.5%: at the algorithm dose
the model delivers exactly the intended 234 ng\*h/mL, which is the
`Dose = CL/F * AUC` identity working as designed.

**The pre-dose leg does not reproduce.** The paper states that an AUC0-12h of
234 ng\*h/mL "corresponds with" a pre-dose concentration of 10 ng/mL, i.e. an
AUC-to-trough ratio of 23.4. The published structural model gives a ratio near
14 instead, so the same 234 ng\*h/mL AUC comes with a trough of about 16.6
ng/mL. The gap is not a steady-state artefact -- the ratio is stable across
the whole post-transplant period:

```{r auc-trough-ratio-over-time}
ratio_over_time <- dplyr::bind_rows(lapply(c(7, 14, 31, 90, 180), function(d) {
  tt <- 24 * d
  ev <- rxode2::et(amt = alg_dose, ii = 12, until = tt, cmt = "depot")
  ev <- rxode2::et(ev, seq(tt, tt + 12, by = 0.05), cmt = "central")
  df <- as.data.frame(ev)
  df$id <- 1L
  for (nm in names(REF)) df[[nm]] <- REF[[nm]]
  r <- rxode2::rxSolve(mod_full_typ, events = df, method = "lsoda") |>
    as.data.frame()
  r <- r[r$time >= tt & !is.na(r$Cc), ]
  r$t <- r$time - tt
  tibble::tibble(
    `Day post-transplant` = d,
    `AUC0-12 (ng*h/mL)`   = round(sum(diff(r$t) * (head(r$Cc, -1) + tail(r$Cc, -1)) / 2), 1),
    `Pre-dose (ng/mL)`    = round(r$Cc[1], 2),
    `AUC / pre-dose`      = round(sum(diff(r$t) * (head(r$Cc, -1) + tail(r$Cc, -1)) / 2) / r$Cc[1], 1)
  )
}))
knitr::kable(ratio_over_time,
             caption = "AUC-to-trough ratio never approaches the 23.4 the 234 ng*h/mL constant implies.")
```

This is recorded as an erratum below; nothing in the model files was tuned to
close the gap.

# The starting-dose model

## Structural check

The same `AUC0-tau = Dose / (CL/F)` identity must hold for the starting-dose
model, at its own typical CL/F of 19.9 L/h.

```{r starting-dose-identity}
subj_start <- tibble::tibble(
  id = 1L, AGE = 57.5, HT = 170,
  CYP3A5_STAR1_HET = 0, CYP3A5_STAR1_HOM = 0,
  SNP_CYP3A4_RS35599367 = 0, STUDY_TACRO_FRANCKE = 3, dose = 5
)

sim_start <- solve_ss(mod_start_typ,
                      make_ss_events(subj_start, obs_times = seq(0, 12, by = 0.05))) |>
  dplyr::filter(!is.na(Cc))

auc_start <- with(sim_start, sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2))
cl_start  <- unique(sim_start$cl)

tibble::tibble(
  Quantity = c("Simulated typical CL/F (L/h)", "Simulated AUC0-12 (ng*h/mL)",
               "Dose / (CL/F) (ng*h/mL)", "Ratio"),
  Value = c(round(cl_start, 3), round(auc_start, 2),
            round(5 / 19.9 * 1000, 2), round(auc_start / (5 / 19.9 * 1000), 5))
) |>
  knitr::kable(caption = "Starting-dose model at 5 mg twice daily, reference covariates.")

stopifnot(length(cl_start) == 1L)
stopifnot(abs(cl_start - 19.9) < 1e-6)
stopifnot(abs(auc_start / (5 / 19.9 * 1000) - 1) < 0.005)
```

## How much does dropping hematocrit and creatinine cost?

Both models describe the same data. At the reference covariate set the two
typical CL/F values differ by only 4%, but the models diverge for recipients
whose hematocrit or creatinine is far from the cohort median -- which is
exactly the information the starting-dose model gives up.

```{r full-vs-starting}
compare_grid <- tidyr::expand_grid(
  HCT = c(0.250, 0.332, 0.420),
  CREAT = c(90, 140, 300)
) |>
  dplyr::mutate(
    `Full model CL/F (L/h)` = cl_of(0, 0, 0, HCT, 57.5, CREAT, 170),
    `Starting-dose CL/F (L/h)` = 19.9,
    `Ratio (full / starting)` = `Full model CL/F (L/h)` / `Starting-dose CL/F (L/h)`
  )

compare_grid |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 3))) |>
  dplyr::rename("Hematocrit (L/L)" = HCT, "Creatinine (umol/L)" = CREAT) |>
  knitr::kable(caption = "The starting-dose model cannot see hematocrit or creatinine, so it predicts one clearance across this whole grid.")
```

## Model-based versus body-weight-based dosing

Section 3.4 compares a standard 0.1 mg/kg twice-daily body-weight dose against
the algorithm dose, reporting that 28.4% versus 33.2% of recipients reach a
pre-dose concentration inside the 7.5-12.5 ng/mL window. The comparison below
uses a 200-recipient virtual cohort drawn from the Table 1 marginals, with
common random numbers across the two dosing arms so the comparison isolates
the dosing rule.

```{r cohort-dosing-comparison}
set.seed(20250630)
N <- 200

cohort <- tibble::tibble(
  id  = seq_len(N),
  AGE = rlnorm_iqr(N, 57.0, 45.0, 65.0),
  HT  = rlnorm_iqr(N, 170, 162, 177),
  WT  = rlnorm_iqr(N, 75.6, 65.0, 86.0),
  HCT = rlnorm_iqr(N, 0.332, 0.300, 0.370),
  CREAT = rlnorm_iqr(N, 140, 110, 191),
  STUDY_TACRO_FRANCKE = 3
) |>
  dplyr::mutate(
    # Genotype frequencies from Table 1, renormalised over genotyped recipients.
    g5 = sample(c("*3/*3", "*1/*3", "*1/*1"), N, replace = TRUE,
                prob = c(884, 224 + 3, 39)),
    CYP3A5_STAR1_HET = as.integer(g5 == "*1/*3"),
    CYP3A5_STAR1_HOM = as.integer(g5 == "*1/*1"),
    SNP_CYP3A4_RS35599367 = as.integer(
      sample(c(0L, 1L), N, replace = TRUE, prob = c(1035, 112))),
    dose_weight = 0.1 * WT,
    dose_model  = mapply(francke_starting_dose, AGE, HT, g5,
                         SNP_CYP3A4_RS35599367 == 1L)
  )

# Common random numbers: one eta draw per recipient, reused in both arms, so
# the comparison isolates the dosing rule rather than sampling noise. Drawn
# with a Cholesky factor of the model's own OMEGA (base R, no extra deps).
omega_full <- mod_full$omega
etas <- matrix(stats::rnorm(N * ncol(omega_full)), nrow = N) %*% chol(omega_full)
colnames(etas) <- dimnames(omega_full)[[1]]
eta_df <- tibble::as_tibble(as.data.frame(etas)) |> dplyr::mutate(id = seq_len(N))

simulate_arm <- function(dose_col, arm_label) {
  subj <- cohort |>
    dplyr::mutate(dose = .data[[dose_col]]) |>
    dplyr::left_join(eta_df, by = "id")
  ev <- make_ss_events(subj, obs_times = c(0, 12))
  out <- rxode2::rxSolve(mod_full, events = ev, method = "lsoda",
                         omega = NA, keep = character()) |>
    as.data.frame()
  if (is.null(out$id)) out$id <- 1L
  out |>
    dplyr::filter(!is.na(Cc), time == 0) |>
    dplyr::transmute(id, arm = arm_label, ctrough = Cc)
}

arms <- dplyr::bind_rows(
  simulate_arm("dose_weight", "Body weight 0.1 mg/kg"),
  simulate_arm("dose_model",  "Starting-dose algorithm")
)

in_window <- arms |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    `Within 7.5-12.5 ng/mL (%)` = round(100 * mean(ctrough >= 7.5 & ctrough <= 12.5), 1),
    `Median pre-dose (ng/mL)`   = round(stats::median(ctrough), 2),
    `IQR width (ng/mL)`         = round(diff(stats::quantile(ctrough, c(0.25, 0.75))), 2),
    .groups = "drop"
  )

knitr::kable(in_window,
             caption = "Simulated pre-dose concentrations, n = 200 per arm, common random numbers.")

ggplot(arms, aes(ctrough, fill = arm)) +
  geom_density(alpha = 0.45) +
  geom_vline(xintercept = c(7.5, 12.5), linetype = "dashed") +
  scale_x_continuous(limits = c(0, 60)) +
  labs(x = "Simulated steady-state pre-dose concentration (ng/mL)",
       y = "Density", fill = "Dosing rule",
       title = "Model-based versus body-weight-based tacrolimus dosing",
       caption = "Compare Francke 2025 Supplementary Figure S9.") +
  theme_minimal()
```

```{r cohort-dosing-assertions}
w_idx <- which(in_window$arm == "Body weight 0.1 mg/kg")
m_idx <- which(in_window$arm == "Starting-dose algorithm")
stopifnot(length(w_idx) == 1L, length(m_idx) == 1L)

# The paper's directional claim: model-based dosing narrows the spread of
# pre-dose concentrations relative to body-weight dosing.
stopifnot(in_window$`IQR width (ng/mL)`[m_idx] < in_window$`IQR width (ng/mL)`[w_idx])
```

The qualitative finding reproduces: the algorithm narrows the interquartile
spread of pre-dose concentrations relative to body-weight dosing. The absolute
percentages inside the 7.5-12.5 ng/mL window are shifted relative to the
paper's 28.4% and 33.2%, which follows directly from the AUC-to-trough
discrepancy documented above -- the algorithm is calibrated so that its
concentrations centre on 10 ng/mL, and under the published structural model
they centre near 16 ng/mL instead.

# Assumptions and deviations

## Extraction decisions

* **Two model files, one vignette.** Francke 2025 reports a base model, a
  final (full) model, and a starting-dose model. The base model is a model
  development step and is not distributed. The full model and the starting-dose
  model are both final, separately validated deliverables with complete
  parameter columns in Table 2 and separate prediction-corrected VPCs
  (Figures 2 and 4), so each is distributed as its own file.
* **IIV scale.** Table 2's `IIV (%)` column is read as %CV, converted with
  `omega^2 = log(1 + CV^2)`. See the scale check above: the %CV reading
  reproduces both the 19.3% and the 35% figures the paper quotes.
* **Correlation matrix.** Table 2 reports eta correlations; the model files
  store covariances as `r_ij * omega_i * omega_j`. Both resulting 3x3 blocks are
  positive definite.
* **Residual error.** Encoded as `lnorm()` with the Table 2 values used
  directly as log-scale SDs, on the authority of the supplement's
  `$SIGMA 1 FIX` construction.
* **Missing covariates.** The control stream maps every missing covariate
  (`-999`) to a multiplier of 1, i.e. to the centring value. The model files
  document this per covariate; a user supplying data must apply the same
  convention (substitute the centring value) rather than leaving `NA`.
* **CYP3A5 "Other" genotype.** The control stream assigns the *1/*3 multiplier
  to both `CYP3A5 == 2` and `CYP3A5 == 4`, so the three recipients with an
  "Other" genotype are coded as heterozygotes.
* **Cohort marginals.** The virtual cohort draws each continuous covariate
  independently from a log-normal matched to the Table 1 median and IQR.
  Francke 2025 does not report a covariate correlation structure, so
  correlations between age, height, weight, hematocrit, and creatinine are
  not reproduced. Genotype frequencies are renormalised over genotyped
  recipients.
* **Figure 3 percentiles.** The paper simulated the 10th, 25th, 50th, 75th,
  and 90th percentiles of each continuous covariate. Table 1 reports only the
  median and IQR, so the panels above use the 25th, 50th, and 75th
  percentiles.
* **Solver.** The default `liblsoda` integrator fails on this model's `ss = 1`
  records; the vignette uses `method = "lsoda"` throughout.

## Errata and inconsistencies in the source

* **Table 2 unit labels.** The `K a` row is headed `(L/h)`; the parameter is a
  first-order absorption rate constant in 1/h, as the Abstract states
  ("the mean absorption rate was 6.59/h"). The `Height` row is headed `(m 2)`;
  the covariate is height in centimetres, as the supplement
  (`CLHGT = ((HGT/170)**THETA(17))`) and the Discussion ("height (in cm)")
  both confirm, and as the centring constant 170 matching the Table 1 median
  height of 170 cm proves.
* **Explained variability quoted on two scales.** The Abstract says the
  covariates "explained 19.3% of the inter-individual variability in
  clearance" while the Key Points, Results, and Conclusions say 35%. Both are
  correct: 19.3% is the reduction on the CV scale and 35% on the variance
  scale. This is not an error, but it is easy to misread.
* **The dosing algorithm mixes two models.** The printed starting-dose formula
  uses the covariate coefficients of the starting-dose model (1.9, 0.842,
  -0.332, 0.97) but the typical CL/F of the *full* model (20.7 L/h); the
  starting-dose model's own CL/F is 19.9 L/h. The formula is reproduced as
  printed, with `cl_constant` exposed as an argument so a user can substitute
  19.9 if the internally consistent version is wanted.
* **The 234 ng\*h/mL AUC target is not consistent with the published
  structural model.** The paper equates an AUC0-12h of 234 ng\*h/mL with a
  10 ng/mL pre-dose concentration (ratio 23.4). Simulating the published
  final model at any time after transplantation gives an AUC-to-trough ratio
  of about 14-15, so 234 ng\*h/mL corresponds to a trough near 16.6 ng/mL and
  a 10 ng/mL trough corresponds to an AUC near 145-155 ng\*h/mL. Section 3.4
  says the constant was determined "on the basis of the data included in this
  study and the population pharmacokinetic model", so the most likely
  explanation is that it was read off the observed AUC/trough pairs in the 208
  densely sampled profiles rather than derived from the model. No parameter
  was adjusted to reconcile the two.
* **`DATA` versus `CENTER`.** The supplement's `$ERROR` block switches the
  residual magnitude on a column called `DATA`, but `$INPUT` lists `CENTER`
  and no `DATA`. The published control stream does not reconcile the two
  names. The model files use a single canonical column,
  `STUDY_TACRO_FRANCKE`, holding the 1-7 stratum code the `$ERROR` block
  expects.
* **Codes 1 vs 2 and 3 vs 4 are undocumented.** The seven centre-by-method
  strata collapse onto five residual magnitudes; the paper never says what
  distinguishes stratum 1 from 2, or 3 from 4. Because each pair shares an
  estimated magnitude, the ambiguity has no effect on predictions.

## Not implemented

* **Inter-occasion variability.** None is reported for either model.
* **The time-after-transplantation effect** (Equation 3) was evaluated by the
  authors and did not improve the model (Section 3.2.2), so it is not encoded.
* **The `V2COV` hook.** The control stream carries `V2COV = 1` with no
  covariates on the central volume in the final model; nothing to encode.
* **Formulation effect on bioavailability.** Prograf versus Adport was tested
  and estimated at exactly 1.0 after correcting for centre (Section 3.2.2), so
  formulation is not a covariate.

# Session info

```{r session-info}
sessionInfo()
```
