---
title: "Dolutegravir maternal-fetal pregnancy PBPK (Freriksen 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dolutegravir maternal-fetal pregnancy PBPK (Freriksen 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

```{r libraries, message = FALSE}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Freriksen et al. (2020, *Clin Pharmacol Ther* 107:1352-1361,
[doi:10.1002/cpt.1748](https://doi.org/10.1002/cpt.1748)) built a
pregnancy physiologically based pharmacokinetic (p-PBPK) model for the
HIV integrase inhibitor **dolutegravir** in Berkeley Madonna, and extended
it with a fetoplacental unit whose placental transfer clearances were
measured in ex vivo dual-side human cotyledon perfusions. The model
simulates maternal plasma, fetal plasma and amniotic-fluid concentrations
after 50 mg once daily at 34 weeks of gestation.

The authors deposited the complete Berkeley Madonna script as
Supplementary File S2, together with the compound table (Table S1) and the
gestational-age regressions (Table S2). This model file is a line-by-line
port of that script.

- Maternal body: 15 perfusion-rate-limited states (oral depot = gut lumen,
  gut wall, lung, adipose, bone, brain, heart, kidney, muscle, skin, spleen,
  liver, rest of body, arterial and venous blood). Kp values were predicted
  by Rodgers-Rowland in Simcyp v17 and are printed in Table S1.
- Liver: well-stirred, with UGT1A1 + CYP3A4 intrinsic clearance scaled up
  from recombinant-enzyme data and multiplied by an empirical scaling
  factor SF = 65. The authors fitted SF against the Min 2010
  healthy-volunteer single-ascending-dose data.
- Placenta: a barrier, not a compartment. The apparent cotyledon clearance
  (1.03 mL/min in each direction) is converted to an unbound clearance with
  the perfusion-buffer fraction unbound, then scaled back to in vivo using
  the blood fraction unbound on each side and 30 cotyledons per placenta.
- Fetus: fetal blood, rest of fetal body, and amniotic fluid. Amniotic-fluid
  exchange uses literature fluid flows scaled by a factor A_SF = 3.

```{r load-model}
mod <- readModelDb("Freriksen_2020_dolutegravir_pregnancy_pbpk")
mod_ui <- rxode2::rxode2(mod)
```

# Population

The model describes one typical pregnant woman at 34 weeks of gestation,
not a population. The authors matched covariates to the mean values of
the clinical studies (Methods, "Physiologically-based pharmacokinetic
modeling"). Her Table S2 regressions give a body weight of 73.7 kg and a
cardiac output of 400.4 L/h at 34 weeks. The simulations were compared
with:

- mean third-trimester maternal profiles from the PANNA network (n = 15)
  and IMPAACT P1026s (n = 28);
- cord-to-maternal plasma ratios from PANNA (n = 10), Rimawi et al.
  (n = 3) and IMPAACT (n = 23);
- one published amniotic-fluid concentration (0.03 ug/mL).

The paper has no between-subject variability and no residual-error model,
so the model file has none either.

# Source trace

| Element | Value | Source |
| --- | --- | --- |
| ODE system (all 18 states) | -- | Supplementary File S2, 'DIFFERENTIAL EQUATIONS'; Methods equations; Figure 1 |
| `ka`, `Fabs` | 2.25 1/h, 0.804 | Table S1 'Absorption'; File S2 |
| 12 maternal Kp + fetal rest-of-body Kp | 0.0730-0.3195 | Table S1 'Distribution'; File S2 |
| `fu` (pregnant), `fu_fet` | 0.0049, 0.0036 | Table S1 'Fraction unbound in plasma' |
| `bpr`, `bpr_fet` | 0.535, 0.441 | Table S1 'Blood:plasma ratio' |
| `CLint` UGT1A1, CYP3A4 | 2.7, 0.56 uL/min/mg | Table S1 'Metabolism/elimination' |
| `fumic`, `MPPGL` | 0.817, 39.79067 mg/g | Table S1; File S2 |
| Liver scaling factor SF | 65 | Table S1; Methods ('refined well-stirred model') |
| CYP3A4 / UGT1A1 induction | 100 + 2.9826 GA - 0.0741 GA^2 (%) | Table S2; File S2 (UGT1A1 assumed equal to CYP3A4) |
| Cotyledon clearances, buffer fu, Ncot | 1.03 / 1.03 mL/min, 0.048 / 0.043, 30 | Table 1; File S2 |
| Placental scaling equation | CL_plac = CL_app / fu_perf * fu_b * Ncot | Methods equations; File S2 |
| Maternal BW, CO | 61.1 + 0.24098 GA + 0.0038 GA^2; 301 + 5.916 GA - 0.088 GA^2 | Table S2; File S2 |
| Fractional volumes and flows | per organ | File S2 ('Gaohua 2012', 'Simcyp v17 healthy female') |
| Fetal volume, fetal blood, amniotic volume | Gompertz, linear, quintic in GA | Table S2; File S2 |
| Fetal cardiac output, fraction to placenta | 553 * VtotalF * 0.06; exp(...) / 100 | Table S2; File S2 |
| Amniotic exchange flows and A_SF | 0.275 / 0.774 / 0.350 L/day; A_SF = 3 | File S2 (Underwood 2005) |
| Validation targets | maternal C0 0.99, fetal Ctrough 1.50 ug/mL, C:M 0.57-1.51, amniotic 0.025-0.039 ug/mL | Results, 'PBPK model in pregnant subjects' |

# Simulation: 50 mg once daily at 34 weeks (Figure 4)

The deposited script gives seven 50 mg doses at 24 h intervals and
simulates to 168 h. The same regimen is used here. Observation rows are
placed on the venous-blood state, and rxode2 returns the algebraic
outputs `Cc` (maternal venous plasma), `Cplasma_fet` and `Camniotic` at
each row.

```{r simulate}
ev <- rxode2::et(amt = 50, cmt = "depot", ii = 24, addl = 6) |>
  rxode2::et(seq(0, 168, by = 0.05), cmt = "venous") |>
  as.data.frame()
ev$GA <- 34
sim <- as.data.frame(rxode2::rxSolve(mod_ui, ev, returnType = "data.frame"))
```

```{r figure4, fig.cap = "Replicates Figure 4 of Freriksen 2020: maternal plasma, fetal plasma and amniotic-fluid dolutegravir after 50 mg once daily at 34 weeks of gestation."}
sim |>
  select(time, `Maternal plasma` = Cc, `Fetal plasma` = Cplasma_fet, `Amniotic fluid` = Camniotic) |>
  pivot_longer(-time, names_to = "matrix", values_to = "conc") |>
  ggplot(aes(time, conc, linetype = matrix, colour = matrix)) +
  geom_line() +
  scale_colour_manual(values = c("Maternal plasma" = "black", "Fetal plasma" = "grey55", "Amniotic fluid" = "black")) +
  scale_linetype_manual(values = c("Maternal plasma" = "solid", "Fetal plasma" = "solid", "Amniotic fluid" = "dotted")) +
  labs(x = "Time (h)", y = "Dolutegravir (ug/mL)", colour = NULL, linetype = NULL) +
  theme_bw()
```

# Comparison with the published numbers

The Results text gives the following deterministic values for the last dosing
interval (144-168 h). All of them come from the same single-subject
solve.

```{r published-checks}
ss <- sim |> filter(time >= 144) |> mutate(tad = time - 144, ratio = Cplasma_fet / Cc)
trough <- ss |> filter(abs(tad - 24) < 1e-6)
hourly <- ss |> filter(abs(tad - round(tad)) < 1e-6)

checks <- data.frame(
  Quantity = c(
    "Maternal plasma C0 (ug/mL)",
    "Fetal plasma Ctrough (ug/mL)",
    "Fetal:maternal plasma ratio, max over interval",
    "Fetal:maternal plasma ratio, min over interval (1 h grid)",
    "Amniotic fluid, min over interval (ug/mL)",
    "Amniotic fluid, max over interval (ug/mL)"
  ),
  Published = c(0.99, 1.50, 1.51, 0.57, 0.025, 0.039),
  Simulated = c(
    trough$Cc, trough$Cplasma_fet, max(ss$ratio), min(hourly$ratio),
    min(ss$Camniotic), max(ss$Camniotic)
  )
) |>
  mutate(`% diff` = round(100 * (Simulated / Published - 1), 1), Simulated = signif(Simulated, 3))
knitr::kable(checks)
```

```{r published-gates}
stopifnot(
  # Maternal and fetal troughs: a mis-transcribed SF, fu, Kp or placental
  # clearance moves these by tens of percent (SF 65 -> 60 alone gives +17%).
  abs(trough$Cc / 0.99 - 1) < 0.10,
  abs(trough$Cplasma_fet / 1.50 - 1) < 0.10,
  # The fetal:maternal ratio depends only on the placental layer.
  abs(max(ss$ratio) / 1.51 - 1) < 0.02,
  # Amniotic-fluid range.
  abs(min(ss$Camniotic) - 0.025) < 0.002,
  abs(max(ss$Camniotic) - 0.039) < 0.002
)
```

The simulated maternal and fetal troughs at 24 h after the last dose are
6% below the published 0.99 and 1.50 ug/mL. One hour earlier (23 h after
the dose) the model gives
`r signif(hourly$Cc[hourly$tad == 23], 3)` and
`r signif(hourly$Cplasma_fet[hourly$tad == 23], 3)` ug/mL. Both round
to the published values, and the ratio between them (1.51) matches the
paper. The uniform 6% offset in both matrices therefore points to the
maternal disposition layer, or to when the published trough was read off
the output, and not to the placental transfer. The trough is very
sensitive to SF: changing it from 65 to 60 raises the trough by 17%. A 6%
offset is within the rounding of the printed inputs, so no parameter was
adjusted.

The smallest fetal:maternal ratio (just after the maternal peak) is 0.54
on the script's 1 h output grid (`DTOUT = 1`), against a published 0.57.
The fetal peak lags the maternal peak, so this minimum depends on how
finely the output is sampled.

# Mass balance

At steady state, the drug eliminated by the liver over one dosing
interval must equal the bioavailable dose, 50 mg x 0.804 = 40.2 mg. The
paper has no other elimination route. The same solve gives the
integrand `cl_liver * cu_liver`. Integrating it over the last interval
checks the ODE topology directly (every flow in must have a matching flow
out).

```{r mass-balance}
elim <- ss |>
  mutate(rate = cl_liver * cu_liver) |>
  summarise(amount = sum(diff(tad) * (head(rate, -1) + tail(rate, -1)) / 2))
elim$amount
stopifnot(abs(elim$amount / (50 * 0.804) - 1) < 0.01)
```

# PKNCA: steady-state interval (day 7)

NCA is run on the last dosing interval, with time re-anchored to the
start of the interval. Two separate NCA blocks are used, one for
maternal plasma and one for fetal plasma.

```{r pknca}
nca_conc <- ss |>
  select(tad, Cc, Cplasma_fet) |>
  pivot_longer(c(Cc, Cplasma_fet), names_to = "matrix", values_to = "conc") |>
  mutate(id = 1L, treatment = "50 mg QD, GA 34 wk") |>
  filter(!is.na(conc))
nca_dose <- data.frame(id = 1L, treatment = "50 mg QD, GA 34 wk", tad = 0, amt = 50)
intervals <- data.frame(start = 0, end = 24, cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, cav = TRUE)

run_nca <- function(m) {
  o_conc <- PKNCA::PKNCAconc(filter(nca_conc, matrix == m), conc ~ tad | treatment + id)
  o_dose <- PKNCA::PKNCAdose(nca_dose, amt ~ tad | treatment + id)
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals))
  as.data.frame(res$result) |> mutate(matrix = m)
}
nca_res <- bind_rows(run_nca("Cc"), run_nca("Cplasma_fet"))
nca_res |>
  select(matrix, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  rename(
    "Matrix" = matrix, "Cmax (ug/mL)" = cmax, "Tmax (h)" = tmax, "Cmin (ug/mL)" = cmin,
    "AUC0-24 (ug*h/mL)" = auclast, "Cav (ug/mL)" = cav
  ) |>
  knitr::kable()
```

The paper reports only the troughs, so the comparison table is limited to
`cmin`. That is the trough (C0) for both matrices, because at steady
state the fetal minimum also falls at the time of dosing.

```{r nca-compare}
sim_cmp <- nca_res |>
  filter(PPTESTCD == "cmin") |>
  select(matrix, PPTESTCD, PPORRES)
ref_cmp <- data.frame(matrix = c("Cc", "Cplasma_fet"), cmin = c(0.99, 1.50))
tbl <- nlmixr2lib::ncaComparisonTable(sim_cmp, ref_cmp, by = "matrix", units = c(cmin = "ug/mL"))
knitr::kable(tbl)
```

# Gestational age

Gestational age (`GA`) drives every physiological regression of Table S2,
so the model accepts it as a covariate. The paper simulated and validated
34 weeks only. The protein-binding inputs `fu` and `fu_fet` are the
34-week values and do not change with GA. The table below is therefore an
extrapolation that shows the physiological drift only.

```{r ga-sweep}
ga_tab <- lapply(c(30, 34, 38), function(g) {
  e <- ev
  e$GA <- g
  s <- as.data.frame(rxode2::rxSolve(mod_ui, e, returnType = "data.frame"))
  s |>
    filter(abs(time - 168) < 1e-6) |>
    transmute(`GA (weeks)` = g, `Maternal C0` = signif(Cc, 3), `Fetal Ctrough` = signif(Cplasma_fet, 3))
}) |> bind_rows()
knitr::kable(ga_tab)
```

# Assumptions and deviations

- **Deposited script over prose.** Every value and equation is taken from
  Supplementary File S2 (Berkeley Madonna script) and Tables 1, S1 and S2.
  Berkeley Madonna identifiers are case-insensitive, so the script's
  `Clplac_mf` and `CLplac_mf` (and `Ka` / `ka`) are the same variable.
  The script's en-dash in the liver equation is a minus sign.
- **Kp scalar.** The script multiplies each maternal Kp by `KpScalar = 1`.
  The factor is omitted here because it is 1.
- **Enzyme induction.** The script applies the same gestational-age factor
  to CYP3A4 and UGT1A1 (the paper assumes UGT1A1 induction is
  proportional to CYP3A4). The model applies it once to the summed
  intrinsic clearance, which is algebraically identical.
- **Hepatic unbound concentration.** As in the script, the unbound liver
  concentration is `C_liver * fu` (plasma fraction unbound), not
  `C_liver / Kp * fu`.
- **Output site.** Maternal plasma is venous plasma (`Cvenous / BP`,
  script `Cven_plasmaM`). The paper does not say whether Figure 4 shows
  venous or arterial plasma. At trough the two agree to 3 significant
  figures.
- **Doses.** The script's `pulse(dose*Fabs, 0, 24)` impulse into the gut
  lumen is encoded as 50 mg bolus doses into `depot` with
  `f(depot) = Fabs`.
- **Trough offset.** The simulated maternal and fetal troughs are 6%
  below the published values (see above). No parameter was tuned.
- **Healthy-volunteer model (Figure 3) not reproduced.** The nonpregnant
  model used to fit SF has a different fraction unbound (0.0041) and
  nonpregnant physiology. Its body weight and cardiac output are not
  given, and the fetal regressions are undefined at GA = 0, so Figure 3
  cannot be reproduced from this file.
- **Third trimester only.** The fetal-blood regression is negative below
  about 18.7 weeks, so the model is not valid early in pregnancy.
- **Specimen coding.** The amniotic-fluid state is coded as specimen
  `tissue`, following other gestational PBPK models in the package,
  because the specimen vocabulary has no amniotic-fluid entry.
- **No variability.** The paper is a single typical-subject forward
  simulation, with no IIV and no residual-error model.
