---
title: "Mycophenolic acid (Fromage 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mycophenolic acid (Fromage 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Fromage_2025_mycophenolic_acid"))$reference`
* Description: `r rxode2::rxode(readModelDb("Fromage_2025_mycophenolic_acid"))$description`
* Article: <https://doi.org/10.1111/bcp.16374>

Mycophenolic acid (MPA) is the active moiety of enteric-coated mycophenolate
sodium (EC-MPS, Myfortic). Fromage 2025 is the first population PK model for
EC-MPS built across multiple indications rather than in renal transplantation
alone, and the first to apply a double-gamma absorption model to this
formulation. The enteric coating delays release into the small intestine and
produces the erratic, frequently double-peaked absorption profiles that
motivate the gamma mixture.

# Population

The model was built from retrospective real-world therapeutic-drug-monitoring
data extracted from the ISBA / ABIS 3.0 Bayesian dose-adaptation platform at
Limoges University Hospital: 153 PK profiles comprising 863 MPA concentrations
from 129 patients (Section 3.1). All patients received Myfortic. Indications
span renal, hepatic, cardiac and pulmonary transplantation, bone marrow /
haematopoietic cell transplantation, and autoimmune disease (systemic lupus
erythematosus, lupus nephritis, nephrotic syndrome): 116 patients with
solid-organ or haematopoietic cell transplantation and 13 with autoimmune
disease. Median age was 45 years (range 6 to 80), and 20 of 127 patients were
paediatric (< 18 years). EC-MPS doses ranged from 180 to 1440 mg per intake
with a median of 360 mg. Sampling was sparse and irregular (3 to 11 samples per
profile, median 5). Baseline characteristics are given in Table 1.

Body weight, serum albumin and creatinine clearance -- the covariates most
often reported to influence MPA clearance -- were **not available** in the ISBA
extraction, which the authors identify as a limitation (Section 4).

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Fromage_2025_mycophenolic_acid")()$population`).

# Model structure

A one-compartment model with first-order elimination and a **double-gamma
absorption** input best described the data (Section 3.2.1). The absorption rate
is a two-component gamma mixture (Section 2.2.1),

$$V_a(t) = F \cdot D \cdot \left[ r \cdot f_1(t) + (1-r) \cdot f_2(t) \right],
\qquad
f_i(t) = \frac{b_i^{a_i}}{\Gamma(a_i)} \, t^{a_i - 1} e^{-b_i t}$$

and the authors give the resulting plasma concentration in closed form,

$$C(t) = C_0 + \mathrm{FAIV} \cdot D \cdot
\left(
r \left(\tfrac{b_1}{b_1 - \lambda}\right)^{a_1} P[a_1, (b_1-\lambda)t]
+ (1-r) \left(\tfrac{b_2}{b_2 - \lambda}\right)^{a_2} P[a_2, (b_2-\lambda)t]
\right) e^{-\lambda t}$$

where $P$ is the regularised lower incomplete gamma function, $C_0$ is the
steady-state trough concentration, $\mathrm{FAIV}$ is the disposition
coefficient and $\lambda$ the disposition rate constant.

That closed form is what Monolix fitted, and it is what the packaged model
evaluates, using rxode2's `gammap(a, z)` for $P[a,z]$. Two consequences are
worth stating up front:

* The gamma components are stored in the canonical transit parameterisation
  `(ntr, mtt)` with $a_i = \mathrm{ntr}_i + 1$ and $b_i = a_i/\mathrm{mtt}_i$,
  so `mtt`$_i$ is exactly the paper's mean absorption time MAT$_i$.
* Because `tad()` and `podo()` refer to the most recent dose, the model
  describes **one steady-state dosing interval** and does not accumulate across
  doses. This matches the authors' framing, in which $C_0$ *is* the trough
  carried in from previous doses.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Fromage_2025_mycophenolic_acid.R`. The
table below collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Absorption input rate `Va(t)` (gamma mixture) | n/a | Section 2.2.1, first two display equations |
| Closed-form `C(t)` | n/a | Section 2.2.1, third display equation |
| `MATi = ai / bi`, `V/F = 100 / FAIV`, `CL/F = Dose / AUC` | n/a | Section 2.2.1, final three display equations |
| Categorical covariate model on `log(b2)` | n/a | Section 2.2.2, categorical-covariate equation and the `log(b2)` IIV equation |
| `lntr1` (a1 = 1280.97) | 1279.97 | Table 2, `a1` (RSE 0.25%) |
| `lmtt1` (a1/b1) | 3.061 h | Table 2, `a1` = 1280.97 and `b1` = 418.42 1/h |
| `lntr2` (a2 = 395.34) | 394.34 | Table 2, `a2` (RSE 2.30%) |
| `lmtt2` (a2/b2) | 4.131 h | Table 2, `a2` = 395.34 and `b2` = 95.69 1/h |
| `lfdepot` (r) | 0.79 | Table 2, `r` (RSE 2.14%) |
| `lvc` (100/FAIV) | 12.71 L | Table 2, `FAIV` = 7.87 mg/L (RSE 3.35%) |
| `lkel` (lambda) | 1.81 1/h | Table 2, `lambda` (RSE 7.48%) |
| `lrbase` (C0) | 2.21 mg/L | Table 2, `C0` (RSE 6.50%) |
| `e_tx_heart_lung_hct_mtt2` | -0.59 | Table 2, `beta_b2_cardiac, pulmonary, bone marrow transplantations` = +0.59 on `log(b2)` (RSE 47.1%); sign inverts on `log(MAT2)` |
| `e_dis_autoimmune_mtt2` | +0.17 | Table 2, `beta_b2_autoimmune diseases` = -0.17 on `log(b2)` (RSE 188%); sign inverts on `log(MAT2)` |
| `etalmtt1` | 0.8464 | Table 2, IIV `b1` = 0.92 SD (RSE 44.9%); variance = SD^2 |
| `etalmtt2` | 0.2704 | Table 2, IIV `b2` = 0.52 SD (RSE 16.1%) |
| `etalvc` | 0.0144 | Table 2, IIV `FAIV` = 0.12 SD (RSE 11.0%) |
| `etalkel` | 0.1936 | Table 2, IIV `lambda` = 0.44 SD (RSE 14.2%) |
| `etalrbase` | 0.2401 | Table 2, IIV `C0` = 0.49 SD (RSE 10.8%) |
| `addSd` | 1.38 mg/L | Table 2, residual additive error `sigma` (RSE 3.47%) |

Table 2's footnote states that "IIV and IOV are expressed as SD", which is why
each `eta` variance above is the reported value squared.

# Gate 1: the packaged model reproduces the published closed form

The strongest available structural check is to evaluate the paper's printed
equation independently -- in plain R, with `stats::pgamma()` supplying
$P[a,z]$ -- and compare it against a solve of the packaged model. Nothing is
fitted or tuned here; agreement to numerical precision means the ODE wiring,
the `(ntr, mtt)` to $(a,b)$ mapping, the `FAIV` to `vc` conversion, the
bioavailability split and the trough offset are all correct simultaneously.

```{r gate1}
mod <- readModelDb("Fromage_2025_mycophenolic_acid")
mod_typ <- rxode2::zeroRe(rxode2::rxode(mod))

# Table 2 values, transcribed once for the independent evaluation.
p <- list(C0 = 2.21, a1 = 1280.97, b1 = 418.42, a2 = 395.34, b2 = 95.69,
          FAIV = 7.87, lambda = 1.81, r = 0.79)
p$vc <- 100 / p$FAIV

published_closed_form <- function(t, dose, b2 = p$b2) {
  arm1 <- p$r * (p$b1 / (p$b1 - p$lambda))^p$a1 *
    stats::pgamma((p$b1 - p$lambda) * t, shape = p$a1)
  arm2 <- (1 - p$r) * (b2 / (b2 - p$lambda))^p$a2 *
    stats::pgamma((b2 - p$lambda) * t, shape = p$a2)
  p$C0 + (dose / p$vc) * (arm1 + arm2) * exp(-p$lambda * t)
}

# Reference indication group (renal / hepatic): all indicators zero.
ref_cov <- list(TX_HEART = 0, TX_LUNG = 0, TX_HCT = 0, DIS_AUTOIMMUNE = 0)

make_events <- function(dose, cov, times) {
  ev <- as.data.frame(rxode2::et(amt = dose, cmt = "depot") |> rxode2::et(times))
  for (nm in names(cov)) ev[[nm]] <- cov[[nm]]
  ev
}

times_fine <- seq(0, 12, by = 0.01)
sim_typ <- rxode2::rxSolve(mod_typ, make_events(720, ref_cov, times_fine))

err <- max(abs(sim_typ$Cc - published_closed_form(sim_typ$time, 720)))
cat(sprintf("max |packaged - published closed form| = %.3e mg/L\n", err))
stopifnot(err < 1e-8)
```

The agreement is at the level of double-precision round-off, so the packaged
model *is* the published equation rather than an approximation of it.

# Gate 2: derived-parameter identities

The paper prints three derived-parameter relations (Section 2.2.1) and quotes
two of the resulting values in the Abstract. Each is checked below.

```{r gate2}
mat1 <- p$a1 / p$b1
mat2 <- p$a2 / p$b2
vdf  <- 100 / p$FAIV

identities <- tibble::tibble(
  Quantity = c("V/F = 100/FAIV (L)", "MAT1 = a1/b1 (h)", "MAT2 = a2/b2 (h)",
               "C(0) = C0, the steady-state trough (mg/L)"),
  Recovered = c(vdf, mat1, mat2, sim_typ$Cc[sim_typ$time == 0]),
  Published = c(12.60, NA_real_, 4.09, 2.21),
  `Source` = c("Abstract (Vd/F = 12.60 L)", "not quoted separately",
               "Discussion (4.09 h; see Errata)", "Table 2 (C0)")
)
knitr::kable(identities, digits = 4,
             caption = "Derived-parameter identities against the published values.")

# The trough identity is exact by construction and must hold.
stopifnot(abs(sim_typ$Cc[sim_typ$time == 0] - p$C0) < 1e-10)
# V/F must land within 1% of the Abstract's rounded 12.60 L.
stopifnot(abs(vdf - 12.60) / 12.60 < 0.01)
```

`V/F` recovers 12.71 L against the Abstract's 12.60 L, a 0.9% difference
attributable to rounding of `FAIV` to three significant figures. The MAT2
discrepancy is a slip in the paper and is analysed in the Errata below.

# Gate 3: the indication covariate reproduces Table 2 and the Discussion

Fromage 2025 retained a single covariate: the three-level indication grouping
acting on the second gamma rate constant `b2`. The Discussion quotes the
resulting `b2` values and mean absorption times for all three groups, which
makes this an unusually well-specified check -- the packaged model must
reproduce six published numbers from two estimated coefficients.

```{r gate3}
groups <- tibble::tribble(
  ~group,                              ~TX_HEART, ~TX_LUNG, ~TX_HCT, ~DIS_AUTOIMMUNE,
  "(i) renal / hepatic (reference)",           0,        0,       0,               0,
  "(ii) cardiac",                              1,        0,       0,               0,
  "(ii) pulmonary",                            0,        1,       0,               0,
  "(ii) haematopoietic cell",                  0,        0,       1,               0,
  "(iii) autoimmune disease",                  0,        0,       0,               1
)

recovered <- lapply(seq_len(nrow(groups)), function(i) {
  cov <- as.list(groups[i, c("TX_HEART", "TX_LUNG", "TX_HCT", "DIS_AUTOIMMUNE")])
  s <- rxode2::rxSolve(mod_typ, make_events(720, cov, times_fine))
  auc <- sum(diff(s$time) * (head(s$Cc, -1) + tail(s$Cc, -1)) / 2)
  tibble::tibble(group = groups$group[i], b2 = s$gb2[1], MAT2 = s$ga2[1] / s$gb2[1],
                 Cmax = max(s$Cc), Tmax = s$time[which.max(s$Cc)], AUC = auc)
}) |> dplyr::bind_rows()

recovered |>
  dplyr::mutate(
    `b2 published` = c(95.69, 172.62, 172.62, 172.62, 80.73),
    `MAT2 published` = c(4.09, 2.29, 2.29, 2.29, 4.90)
  ) |>
  dplyr::rename("Indication group" = group, "b2 (1/h)" = b2, "MAT2 (h)" = MAT2,
                "Cmax (mg/L)" = Cmax, "Tmax (h)" = Tmax, "AUC0-12 (mg*h/L)" = AUC) |>
  knitr::kable(digits = c(0, 2, 2, 3, 2, 2, 2, 2),
               caption = paste("Indication effect on the second gamma component.",
                               "Published b2 values and MAT2 are from the Discussion."))

# The three pooled group-(ii) indicators must give an identical b2, because the
# source estimates ONE effect for the pooled level.
grp2 <- recovered$b2[recovered$group %in%
                       c("(ii) cardiac", "(ii) pulmonary", "(ii) haematopoietic cell")]
stopifnot(length(grp2) == 3, diff(range(grp2)) < 1e-9)

# b2 must reproduce the Discussion's printed values to rounding.
stopifnot(max(abs(recovered$b2 - c(95.69, 172.62, 172.62, 172.62, 80.73))) < 0.01)

# AUC0-12 must be INVARIANT across indication groups: b2 shifts the timing of
# the second absorption peak but not the amount absorbed. This is a genuine
# falsifier -- had the covariate been wired onto vc or kel instead of mtt2,
# these AUCs would differ.
stopifnot(diff(range(recovered$AUC)) / mean(recovered$AUC) < 1e-6)
```

All five rows reproduce the published `b2` exactly, the three pooled group-(ii)
indicators are numerically identical as they must be, and AUC0-12 is invariant
across groups -- confirming the covariate acts on absorption *timing* only.
Group (ii) roughly halves MAT2 (4.13 h to 2.29 h), which is the effect the
authors describe.

# Virtual cohort

Original observed data are not publicly available. The cohort below uses the
two most common EC-MPS strengths in the source (the median 360 mg intake and a
720 mg intake), 200 subjects per arm, with the reference renal / hepatic
indication so the arms differ only in dose.

```{r cohort}
set.seed(20250421)

n_per_arm <- 200L
arm_doses <- c(`360 mg` = 360, `720 mg` = 720)

# Observation grid: 0.05 h over the interval, refined to 0.01 h across the
# absorption peak so Cmax and Tmax are resolved. The peak is genuinely sharp
# (component 1 has width ~sqrt(a1)/b1 = 0.086 h).
obs_times <- sort(unique(c(seq(0, 12, by = 0.05), seq(2.0, 4.5, by = 0.01))))

make_arm <- function(dose, label, id_offset) {
  ids <- id_offset + seq_len(n_per_arm)
  dosing <- tibble::tibble(id = ids, time = 0, amt = dose, evid = 1L,
                           cmt = "depot")
  obs <- tidyr::expand_grid(id = ids, time = obs_times) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "depot")
  dplyr::bind_rows(dosing, obs) |>
    dplyr::mutate(arm = label, TX_HEART = 0, TX_LUNG = 0, TX_HCT = 0,
                  DIS_AUTOIMMUNE = 0) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  make_arm(arm_doses[["360 mg"]], "360 mg", id_offset = 0L),
  make_arm(arm_doses[["720 mg"]], "720 mg", id_offset = n_per_arm)
)

# Disjoint-ID guard: each subject must belong to exactly one arm. Duplicate IDs
# across arms silently merge into one subject receiving the summed dose.
ids_per_arm <- events |> dplyr::distinct(id, arm) |> dplyr::count(id)
stopifnot(nrow(ids_per_arm) == 2L * n_per_arm, all(ids_per_arm$n == 1L))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(rxode2::rxode(mod), events = events, keep = c("arm")) |>
  as.data.frame()

# The solve must produce usable concentrations for every subject: a silent
# solver failure in this model surfaces as Cc collapsing to the trough offset
# (see Errata), so assert that absorption actually happened in both arms.
peak_by_arm <- sim |> dplyr::group_by(arm) |>
  dplyr::summarise(peak = max(Cc, na.rm = TRUE), .groups = "drop")
stopifnot(!anyNA(sim$Cc), all(peak_by_arm$peak > 3 * p$C0))
```

# Replicate published figures

```{r figure-2}
# Replicates Figure 2 of Fromage 2025: individual MPA concentration-time
# profiles showing the double-peaked absorption the gamma mixture is designed
# to capture. Nine representative simulated subjects from the 720 mg arm.
sim |>
  dplyr::filter(arm == "720 mg", id %in% (n_per_arm + 1:9)) |>
  ggplot(aes(time, Cc, group = id)) +
  geom_line(colour = "#6a3d9a", linewidth = 0.6) +
  facet_wrap(~id, ncol = 3, scales = "free_y") +
  labs(x = "Time (h)", y = "MPA concentration (mg/L)",
       title = "Figure 2 -- representative individual profiles (720 mg)",
       caption = "Replicates Figure 2 of Fromage 2025.") +
  theme_bw()
```

```{r figure-4}
# Replicates Figure 4 of Fromage 2025: prediction-corrected VPC. Shown here as
# the simulated 10th / 50th / 90th percentiles by dose arm (the source overlays
# observed data that are not publicly available).
sim |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(Q10 = quantile(Cc, 0.10), Q50 = quantile(Cc, 0.50),
                   Q90 = quantile(Cc, 0.90), .groups = "drop") |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.25, fill = "#1f78b4") +
  geom_line(colour = "#1f78b4", linewidth = 0.7) +
  facet_wrap(~arm) +
  labs(x = "Time (h)", y = "MPA concentration (mg/L)",
       title = "Figure 4 -- simulated prediction interval by dose arm",
       caption = paste("Replicates the prediction-interval component of",
                       "Figure 4 of Fromage 2025 (10th to 90th percentile).")) +
  theme_bw()
```

The simulated profiles show the shoulder-then-peak shape the double-gamma model
was chosen for, and the wide prediction interval reflects the large IIV on the
absorption rate constants (SD 0.92 on `log(b1)`).

# PKNCA validation

NCA is computed over the 0 to 12 h steady-state dosing interval. `cmin` is used
rather than `ctrough` because `ctrough` is all-`NA` on a steady-state interval
of this shape.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

# Guarantee a time = 0 row per (id, arm) so AUC0-12 is anchored. For this model
# the time-zero concentration is the trough C0, not zero, so keep the simulated
# value where one exists rather than overwriting it.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = p$C0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, arm, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, arm)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)

intervals <- data.frame(start = 0, end = 12,
                        cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(arm, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

stopifnot(nrow(nca_wide) == 2L * n_per_arm, !anyNA(nca_wide$auclast))

nca_wide |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    `Cmax (mg/L)`       = mean(cmax),
    `Tmax (h)`          = median(tmax),
    `Cmin (mg/L)`       = mean(cmin),
    `AUC0-12 (mg*h/L)`  = mean(auclast),
    .groups = "drop"
  ) |>
  dplyr::rename("Dose arm" = arm) |>
  knitr::kable(digits = 2,
               caption = "Simulated steady-state NCA by dose arm (n = 200 per arm).")
```

## Comparison against published values

Fromage 2025 does not tabulate Cmax, Tmax or AUC, so the comparison targets the
two derived quantities the Abstract does report: `CL/F = 4.99 +/- 2.22 L/h` and
`Vd/F = 12.60 +/- 0.08 L` (mean +/- SD). The paper defines `CL/F = Dose / AUC`
(Section 2.2.1), which is computed below from the PKNCA `auclast` over the
dosing interval.

```{r nca_compare}
clf <- nca_wide |>
  dplyr::mutate(dose = dplyr::if_else(arm == "360 mg", 360, 720),
                clf = dose / auclast)

comparison <- clf |>
  dplyr::group_by(arm) |>
  dplyr::summarise(Simulated = mean(clf), `Simulated SD` = sd(clf),
                   .groups = "drop") |>
  dplyr::mutate(Quantity = paste0("CL/F, ", arm, " (L/h)"),
                Published = 4.99, `Published SD` = 2.22) |>
  dplyr::select(Quantity, Simulated, `Simulated SD`, Published, `Published SD`) |>
  dplyr::bind_rows(tibble::tibble(
    Quantity = "Vd/F (L)", Simulated = 100 / p$FAIV, `Simulated SD` = NA_real_,
    Published = 12.60, `Published SD` = 0.08))

comparison |>
  dplyr::mutate(`Ratio sim/pub` = Simulated / Published) |>
  knitr::kable(digits = 2,
               caption = paste("Simulated vs published derived parameters.",
                               "CL/F is Dose/AUC0-12 per the paper's definition."))
```

`Vd/F` matches to 0.9%. `CL/F` needs interpretation, and the reason is
structural rather than a defect: because `C0` is a **dose-independent** additive
offset, it contributes a fixed `12 * C0 = 26.5 mg*h/L` to every AUC0-12
regardless of dose, so `Dose/AUC` is necessarily dose-dependent under this
model. Evaluating the typical-value profile across the source's full dose range
makes the pattern explicit:

```{r clf-dose-dependence}
dose_scan <- lapply(c(180, 360, 540, 720, 1080, 1440), function(d) {
  s <- rxode2::rxSolve(mod_typ, make_events(d, ref_cov, times_fine))
  auc <- sum(diff(s$time) * (head(s$Cc, -1) + tail(s$Cc, -1)) / 2)
  tibble::tibble(`Dose (mg)` = d, `AUC0-12 (mg*h/L)` = auc,
                 `CL/F = Dose/AUC (L/h)` = d / auc)
}) |> dplyr::bind_rows()

knitr::kable(dose_scan, digits = 2,
             caption = paste("Typical-value CL/F across the source's dose range",
                             "(180 to 1440 mg per intake)."))
```

The typical-value `CL/F` runs from 5.2 L/h at the lowest studied intake
(180 mg) to 16.2 L/h at the highest (1440 mg). The published mean of
4.99 L/h therefore sits at the **low-dose end** of the model's own range. Two
factors plausibly account for this without any parameter being wrong: the
published figure is an arithmetic mean of *individual* `Dose/AUC` values over a
cohort with a right-skewed AUC distribution (large IIV on `C0`, `lambda` and
`b1`), and low-dose occasions -- where `Dose/AUC` is smallest -- are heavily
represented (Table 1 median intake 360 mg, minimum 180 mg). No parameter was
adjusted to close the gap; the discrepancy is reported as found.

# Assumptions and deviations

* **`C0` is a constant offset, not a decaying initial condition.** The closed
  form adds `C0` outside the `exp(-lambda*t)` term, and the paper defines it as
  "the estimated trough concentration at steady state". Reading it as a
  decaying initial condition is falsified twice over: with `lambda = 1.81 1/h`
  (t1/2 = 0.38 h) a decaying `C0` would be indistinguishable from zero by 12 h,
  contradicting the ~2 mg/L troughs in Figure 2; and it would imply `CL/F`
  around 21 to 23 L/h at every plausible dose, against the Abstract's
  4.99 L/h.
* **`FAIV` is normalised per 100 mg.** It is reported in mg/L and the paper
  prints `V/F = 100/FAIV`, so the closed form's `FAIV * D` amplitude is `D/vc`
  with `vc = 100/FAIV = 12.71 L`. This is what reproduces the Abstract's
  `Vd/F = 12.60 L`.
* **Gamma parameters stored as `(ntr, mtt)`.** The bijection
  `a = ntr + 1`, `b = a/mtt` keeps the canonical transit parameterisation used
  by `Debord_2001_cyclosporin.R` (the same group's gamma absorption model)
  while remaining exactly the paper's `(a, b)`. Because `a` carries no IIV, the
  paper's IIV on `log(b)` and its covariate effect on `log(b2)` transfer to
  `log(mtt)` with a **sign flip** and unchanged variance; the sign inversion is
  recorded in the `ini()` comments and verified in Gate 3.
* **Single dosing interval only.** `tad()` / `podo()` refer to the most recent
  dose, so each dose restarts the absorption input and there is no
  superposition. The model must not be used to build up accumulation across
  doses; `C0` already represents the steady-state carry-over.
* **Inter-occasion variability is not encoded.** Table 2 reports IOV on `b1`
  (SD 1.08, RSE 28.5%), estimated from sequential visits. Library model files
  omit IOV by nlmixr2lib convention, since no occasion column is assumed; a
  user who needs it can add an IOV term on `lmtt1` with variance 1.08^2. The
  authors themselves caution that highly variable visit intervals (mean 201
  days, range 2 to 813) may have biased this estimate.
* **The autoimmune-disease effect is retained despite being non-significant.**
  Section 3.2.2 reports `beta_b2_autoimmune` with 188% RSE and states it "can
  be disregarded", but Section 4 confirms the authors "retained it in the model
  used for pcVPC simulations". It is retained here for fidelity to the model
  the authors simulated from. Users wanting the reduced model can set
  `e_dis_autoimmune_mtt2` to 0.
* **Covariates screened but not retained** (sex, age, time since
  transplantation, immunosuppressive co-medication, analytical method) are
  recorded in the model file's `covariatesDataExcluded` list rather than
  `covariateData`, preserving the provenance of the covariate screen without
  declaring covariates that `model()` never references.
* **Body weight, albumin and creatinine clearance are absent.** These are the
  covariates most often reported to affect MPA clearance, but the ISBA
  extraction did not contain them (Section 4), so the model has no weight
  scaling. This limits extrapolation to the paediatric tail of the cohort
  (youngest 6 years) even though such patients contributed data.
* **Virtual cohort composition.** Both arms use the reference renal / hepatic
  indication and differ only in dose, so the dose comparison is not confounded
  by the absorption covariate. Sex, age and race are not model covariates and
  are therefore not simulated.

## Errata

* **MAT2 for the reference group.** The Discussion states that `b2` values of
  95.69, 172.62 and 80.73 1/h give MAT2 of "4.09 h, 2.29 h and 4.90 h"
  respectively. With `a2 = 395.34` from Table 2, `a2/b2` gives
  **4.13 h**, 2.29 h and 4.90 h. The two covariate-group values reproduce
  exactly from the same `a2`, and no single `a2` reproduces all three printed
  values, so 4.09 h is a rounding or transcription slip in the paper rather
  than a different parameter set. The packaged model implies 4.13 h; Gate 2
  reports both numbers.
* **Numerical note on the closed form versus an ODE.** The published closed
  form is used, not an equivalent ODE, partly because it is what was fitted and
  partly for robustness: `a1 = 1280.97` makes the first gamma component a
  near-delta pulse of width ~0.086 h, and integrating the corresponding ODE
  drives the central amount through denormal magnitudes (~1e-181) before the
  pulse arrives. With rxode2 5.1.7 that collapses lsoda's step size, and on
  output grids finer than about 0.02 h the ODE formulation returns no
  absorption at all -- `Cc` flat at the trough offset, with **no error or
  warning** under the default `liblsoda` -- unless `atol` is loosened to 1e-4
  or larger. The closed form has no such failure mode and agrees with an
  independent `pgamma()` evaluation to 7e-15 mg/L at every grid density tested.
  The assertion in the Simulation chunk above guards against a silent
  recurrence.
