---
title: "Vancomycin (Fu 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vancomycin (Fu 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Fu_2021_vancomycin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Fu_2021_vancomycin"))$description`
* Article: <https://doi.org/10.1002/bdd.2303> (open access, PMC9297986)

Fu 2021 reports a two-compartment population PK model of intravenous
vancomycin in patients with haematological diseases and neutropenia, uses
Monte Carlo simulation of it to pick a starting regimen by renal function
(Figure 3), and then tests that starting regimen prospectively (Table 5). The
model itself was first published in Chinese by Lin et al. (Chinese Journal of
Clinical Pharmacology and Therapeutics, 2021); every value used here is
printed in Fu 2021.

# Population

The model was fit to 77 patients (42 male, 35 female) aged 17-83 years
(mean 43.28) at Hainan General Hospital, Haikou, China, between January 2018
and January 2020 (Fu 2021 Results 3.1 and Table 1). All had a haematological
disease -- acute myeloid leukaemia (34), acute lymphoblastic leukaemia (16),
lymphoma (10), aplastic anaemia (5), myelodysplastic syndrome (5), chronic
myeloid leukaemia (4), multiple myeloma (2), chronic monocytic leukaemia (1) --
and neutropenia (ANC < 0.5 x 10^9/L) during intermittent intravenous
vancomycin; patients on any blood purification were excluded. Body weight was
56.84 +/- 11.09 kg (33.00-83.50) and CKD-EPI creatinine clearance
118.78 +/- 22.69 mL/min/1.73 m^2 (45.60-163.80): a lean cohort with high renal
function. 80.74% of regimens were 1 g q12h; infusions ran at about
900 mg/h (Table 2). 109 troughs and 43 peaks were measured.

The same information is available programmatically:

```{r population}
str(rxode2::rxode(readModelDb("Fu_2021_vancomycin"))$population)
```

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Structure: two compartments, IV infusion into `central` | n/a | Fu 2021 Results 3.2.1 |
| `CL = 6.84 * (WT/70)^0.75 * (CRCL/116)^0.895 * exp(eta1)` | n/a | Fu 2021 equation (6) |
| `V1 = 20.5 * (WT/70) * exp(eta2)` | n/a | Fu 2021 equation (7) |
| `Q = 15.2 * (WT/70)^0.75 * exp(eta3)` | n/a | Fu 2021 equation (8) |
| `V2 = 50 * (WT/70) * exp(eta4)` | n/a | Fu 2021 equation (9) |
| `lcl` | log(6.84 L/h) | Table 3 (RSE 4.5%) |
| `lvc` | log(20.5 L) | Table 3 (RSE 17.3%) |
| `lq` | log(15.2 L/h) | Table 3 (RSE 23.4%) |
| `lvp` | log(50 L) | Table 3 (RSE 27.2%) |
| `e_crcl_cl` | 0.895 | Table 3, theta CLCR_CL (RSE 14.3%) |
| `e_wt_cl`, `e_wt_q` | 0.75, held constant | Equations (6) and (8); no estimate in Table 3 |
| `e_wt_vc`, `e_wt_vp` | 1, held constant | Equations (7) and (9); no estimate in Table 3 |
| `etalcl` | 0.178^2 = 0.031684 | Table 3, eta1 = 17.8% (RSE 15%) |
| `etalvc` | 0.33^2 = 0.1089 | Table 3, eta2 = 33% (RSE 35.8%) |
| `etalq`, `etalvp` | 0, held constant | Equations (8)-(9) carry eta3/eta4; no magnitude reported |
| `propSd`, `addSd` | 0, held constant | Combined error (Methods 2.2, Results 3.2.1); no magnitude reported |

# Replicate Figure 3: first trough by renal function and regimen

Fu 2021 Figure 3 shows the simulated vancomycin trough for four regimens
(1 g q12h, 1 g q8h, 0.5 g q6h, 0.5 g q8h) at CKD-EPI creatinine clearance of
30, 60, 90 and 120 mL/min/1.73 m^2. Results 3.3 places the trough at "the 48th
hour (the first C0)", which the clinical arm samples 30 min before the next
dose; the infusions are simulated at 1000 mg/h, the median rate in Table 2.
The paper does not state the body weight of the simulated patients. The
maintainers digitised the box medians from the figure, calibrating the y axis
on its three dotted reference lines at 10, 15 and 20 ug/mL.

```{r fig3-digitised}
fig3_digitised <- tibble::tribble(
  ~CRCL, ~regimen,     ~paper_median,
  30,    "1 g q12h",   23.66,
  30,    "1 g q8h",    37.66,
  30,    "0.5 g q6h",  25.83,
  30,    "0.5 g q8h",  18.82,
  60,    "1 g q12h",   13.25,
  60,    "1 g q8h",    22.31,
  60,    "0.5 g q6h",  15.70,
  60,    "0.5 g q8h",  10.95,
  90,    "1 g q12h",    8.34,
  90,    "1 g q8h",    14.60,
  90,    "0.5 g q6h",  10.29,
  90,    "0.5 g q8h",   7.17,
  120,   "1 g q12h",    5.63,
  120,   "1 g q8h",    10.26,
  120,   "0.5 g q6h",   7.39,
  120,   "0.5 g q8h",   5.03
)

regimens <- tibble::tribble(
  ~regimen,    ~dose, ~tau,
  "1 g q12h",  1000,  12,
  "1 g q8h",   1000,   8,
  "0.5 g q6h",  500,   6,
  "0.5 g q8h",  500,   8
)
stopifnot(setequal(regimens$regimen, fig3_digitised$regimen))
```

```{r helpers}
# Build an event table: 1000 mg/h infusions every `tau` h from time 0 up to,
# but not including, 48 h, and observations on the `central` state.
make_events <- function(subjects, obs_times) {
  doses <- subjects |>
    dplyr::rowwise() |>
    dplyr::reframe(
      id = id,
      time = seq(0, 48 - 1e-6, by = tau),
      amt = dose,
      rate = 1000,
      evid = 1L
    )
  obs <- tidyr::crossing(id = subjects$id, time = obs_times) |>
    dplyr::mutate(amt = NA_real_, rate = NA_real_, evid = 0L)
  dplyr::bind_rows(doses, obs) |>
    dplyr::mutate(cmt = "central") |>
    dplyr::left_join(subjects, by = "id") |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}
```

## Typical-value prediction

```{r fig3-typical}
mod <- readModelDb("Fu_2021_vancomycin")
mod_typical <- rxode2::zeroRe(mod)

grid_typical <- tidyr::crossing(regimens, CRCL = c(30, 60, 90, 120), WT = c(70, 57)) |>
  dplyr::mutate(id = dplyr::row_number())
ev_typical <- make_events(grid_typical |> dplyr::select(id, dose, tau, CRCL, WT), obs_times = 47.5)
stopifnot(!anyDuplicated(ev_typical[ev_typical$evid == 1L, c("id", "time")]))

sim_typical <- rxode2::rxSolve(mod_typical, events = ev_typical) |>
  as.data.frame() |>
  dplyr::filter(time == 47.5) |>
  dplyr::select(id, Cc) |>
  dplyr::left_join(grid_typical, by = "id")

fig3_typical <- sim_typical |>
  dplyr::select(regimen, CRCL, WT, Cc) |>
  tidyr::pivot_wider(names_from = WT, values_from = Cc, names_prefix = "WT_") |>
  dplyr::inner_join(fig3_digitised, by = c("regimen", "CRCL")) |>
  dplyr::mutate(
    pct_70 = 100 * (WT_70 / paper_median - 1),
    pct_57 = 100 * (WT_57 / paper_median - 1)
  ) |>
  dplyr::arrange(CRCL, match(regimen, regimens$regimen))
stopifnot(nrow(fig3_typical) == 16L)

fig3_typical |>
  dplyr::mutate(dplyr::across(c(WT_70, WT_57, pct_70, pct_57), \(x) round(x, 1))) |>
  dplyr::select(CRCL, regimen, paper_median, WT_70, pct_70, WT_57, pct_57) |>
  dplyr::rename(
    "CRCL (mL/min/1.73 m^2)" = CRCL,
    "Regimen" = regimen,
    "Figure 3 median (ug/mL)" = paper_median,
    "Model, 70 kg (ug/mL)" = WT_70,
    "Diff, 70 kg (%)" = pct_70,
    "Model, 57 kg (ug/mL)" = WT_57,
    "Diff, 57 kg (%)" = pct_57
  ) |>
  knitr::kable(caption = "Typical-value 47.5-h trough against the digitised Figure 3 box medians.")
```

At the 70 kg reference weight the typical-value trough lands on every one of
the 16 Figure 3 medians to within about 5%, which is the resolution of the
digitisation. At the cohort's mean weight of 57 kg the predictions are
uniformly about 16-21% higher, so the figure was simulated at the 70 kg
reference. The check is deterministic, so it is gated tightly: a
mis-transcribed clearance, renal exponent or volume, or a wrong trough time,
moves whole rows of this table by tens of percent.

```{r fig3-typical-gate}
stopifnot(
  max(abs(fig3_typical$pct_70)) < 8,
  abs(median(fig3_typical$pct_70)) < 4,
  # The cohort-mean weight does not reproduce the figure, so the reading of
  # the unstated simulation weight is itself checked.
  median(fig3_typical$pct_57) > 10
)
```

## Monte Carlo replicate

```{r fig3-stochastic}
rxode2::rxSetSeed(2021)
n_per_cell <- 200L
grid_mc <- tidyr::crossing(regimens, CRCL = c(30, 60, 90, 120)) |>
  dplyr::mutate(cell = dplyr::row_number())
subjects_mc <- grid_mc |>
  tidyr::uncount(n_per_cell) |>
  dplyr::mutate(id = dplyr::row_number(), WT = 70)
ev_mc <- make_events(subjects_mc |> dplyr::select(id, dose, tau, CRCL, WT), obs_times = 47.5) |>
  dplyr::left_join(subjects_mc |> dplyr::select(id, regimen), by = "id")

sim_mc <- rxode2::rxSolve(mod, events = ev_mc, keep = c("regimen", "CRCL")) |>
  as.data.frame() |>
  dplyr::filter(time == 47.5)

fig3_mc <- sim_mc |>
  dplyr::group_by(regimen, CRCL) |>
  dplyr::summarise(
    q25 = quantile(Cc, 0.25),
    sim_median = median(Cc),
    q75 = quantile(Cc, 0.75),
    .groups = "drop"
  ) |>
  dplyr::inner_join(fig3_digitised, by = c("regimen", "CRCL")) |>
  dplyr::mutate(pct = 100 * (sim_median / paper_median - 1))
stopifnot(nrow(fig3_mc) == 16L)

# Centre and envelope only, never the extremes of a simulated cohort.
stopifnot(
  abs(median(fig3_mc$pct)) < 6,
  quantile(abs(fig3_mc$pct), 0.9) < 12
)

sim_mc |>
  dplyr::mutate(regimen = factor(regimen, levels = regimens$regimen)) |>
  ggplot(aes(factor(CRCL), Cc, fill = regimen)) +
  geom_hline(yintercept = c(10, 15, 20), linetype = "dotted") +
  geom_boxplot(outlier.shape = NA, position = position_dodge(width = 0.85)) +
  geom_point(
    data = fig3_digitised |> dplyr::mutate(regimen = factor(regimen, levels = regimens$regimen)),
    aes(factor(CRCL), paper_median, group = regimen),
    position = position_dodge(width = 0.85), shape = 4, size = 2.5, inherit.aes = FALSE
  ) +
  coord_cartesian(ylim = c(0, 60)) +
  labs(
    x = "CRCL (mL/min/1.73 m^2)", y = "Trough at 47.5 h (ug/mL)", fill = "Regimen",
    title = "First trough by renal function and regimen (WT 70 kg)",
    caption = "Replicates Figure 3 of Fu 2021; crosses are the digitised published medians."
  )
```

# Clinical application arm (Table 5)

Table 5 reports the first trough at 48 h in 74 further patients with
CRCL >= 90 mL/min/1.73 m^2: 32 started on 1 g q8h (the model-guided group) and
42 on 1 g q12h (the empiric group). Each arm is simulated with 200 virtual
patients whose weight and CRCL are drawn from normal distributions with the
Table 5 means and SDs, truncated to the Table 5 ranges.

```{r table5}
rtnorm <- function(n, mean, sd, lo, hi) {
  x <- rnorm(n, mean, sd)
  while (any(out <- x < lo | x > hi)) x[out] <- rnorm(sum(out), mean, sd)
  x
}
set.seed(2021)
n_arm <- 200L
arms <- dplyr::bind_rows(
  tibble::tibble(
    arm = "Model group, 1 g q8h", dose = 1000, tau = 8,
    WT = rtnorm(n_arm, 56.57, 9.40, 38.0, 72.0),
    CRCL = rtnorm(n_arm, 125.46, 17.57, 92.2, 164.8)
  ),
  tibble::tibble(
    arm = "Non-model group, 1 g q12h", dose = 1000, tau = 12,
    WT = rtnorm(n_arm, 56.43, 9.56, 48.0, 73.5),
    CRCL = rtnorm(n_arm, 119.63, 18.62, 93.6, 167.3)
  )
) |>
  dplyr::mutate(id = dplyr::row_number())
ev_t5 <- make_events(arms |> dplyr::select(id, dose, tau, CRCL, WT), obs_times = 47.5) |>
  dplyr::left_join(arms |> dplyr::select(id, arm), by = "id")
stopifnot(!anyDuplicated(unique(ev_t5[, c("id", "time", "evid")])))

rxode2::rxSetSeed(2021)
sim_t5 <- rxode2::rxSolve(mod, events = ev_t5, keep = "arm") |>
  as.data.frame() |>
  dplyr::filter(time == 47.5)

t5 <- sim_t5 |>
  dplyr::group_by(arm) |>
  dplyr::summarise(sim_mean = mean(Cc), sim_sd = sd(Cc), sim_median = median(Cc), .groups = "drop") |>
  dplyr::left_join(
    tibble::tribble(
      ~arm,                         ~obs_mean, ~obs_sd,
      "Model group, 1 g q8h",       14.30,     4.73,
      "Non-model group, 1 g q12h",   8.02,     2.61
    ),
    by = "arm"
  ) |>
  dplyr::mutate(pct = 100 * (sim_mean / obs_mean - 1))
stopifnot(nrow(t5) == 2L)

t5 |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 2))) |>
  dplyr::select(arm, obs_mean, obs_sd, sim_mean, sim_sd, sim_median, pct) |>
  dplyr::rename(
    "Arm" = arm,
    "Observed mean (ug/mL)" = obs_mean,
    "Observed SD" = obs_sd,
    "Simulated mean (ug/mL)" = sim_mean,
    "Simulated SD" = sim_sd,
    "Simulated median" = sim_median,
    "Mean diff (%)" = pct
  ) |>
  knitr::kable(caption = "First trough at 48 h, Fu 2021 Table 5 versus simulation.")

ratio_obs <- 14.30 / 8.02
ratio_sim <- t5$sim_mean[t5$arm == "Model group, 1 g q8h"] /
  t5$sim_mean[t5$arm == "Non-model group, 1 g q12h"]
c(observed = ratio_obs, simulated = ratio_sim)
```

The model under-predicts both observed mean troughs, by roughly 10-15%. The
direction is the same in both arms and the q8h/q12h ratio (observed 1.78) is
reproduced closely, which points to a level difference between this
prospective cohort and the model population rather than to a structural
error. The observed SDs are larger than the simulated ones because the model
carries no residual error (none is reported) and no variability on Q and V2.
The gate checks the ratio, which every arm-level structural error would move,
and records the level difference as a known deviation.

```{r table5-gate}
# Each arm mean has a Monte Carlo SE of about 2% at n = 200, so the ratio
# carries about 3%; the bound sits more than 3 SE beyond the ~6% realised
# here, while a mis-transcribed volume or clearance moves it by far more.
stopifnot(
  abs(ratio_sim / ratio_obs - 1) < 0.20,
  # Known deviation (both arms ~10-15% low): bounded, not widened to zero.
  all(t5$pct > -35 & t5$pct < 10)
)
```

# Worked case (Results 3.5)

A 50-year-old woman of 47 kg with a CKD-EPI CRCL of 128.6 mL/min/1.73 m^2
received 1 g q12h; her first trough at 48 h was 6.0 ug/mL. The paper's
(individualised) prediction for 1 g q8h was 11.8 ug/mL and the measured trough
after the switch 10.2 ug/mL.

```{r case}
case <- tibble::tibble(id = 1:2, dose = 1000, tau = c(12, 8), WT = 47, CRCL = 128.6)
sim_case <- rxode2::rxSolve(mod_typical, events = make_events(case, 47.5)) |>
  as.data.frame() |>
  dplyr::filter(time == 47.5)
tibble::tibble(
  Regimen = c("1 g q12h", "1 g q8h"),
  "Typical-value prediction (ug/mL)" = round(sim_case$Cc, 1),
  "Reported (ug/mL)" = c("6.0 measured", "11.8 predicted, 10.2 measured")
) |>
  knitr::kable(caption = "Typical-value trough at 47.5 h for the Results 3.5 patient.")
```

The typical-value predictions (no individual adaptation) lie 8-25% above the
reported values, as expected for a single patient.

# PKNCA validation

The paper reports no NCA. PKNCA is therefore run on a steady-state dosing
interval of the model-population virtual cohort for the two main regimens,
and the AUC over the interval is checked against its closed form,
`AUC_tau,ss = Dose / CL`, subject by subject.

```{r pknca}
set.seed(20211)
n_nca <- 200L
nca_subjects <- tidyr::crossing(
  tibble::tibble(
    sid = seq_len(n_nca),
    WT = rtnorm(n_nca, 56.84, 11.09, 33.0, 83.5),
    CRCL = rtnorm(n_nca, 118.78, 22.69, 45.6, 163.8)
  ),
  regimens |> dplyr::filter(dose == 1000)
) |>
  dplyr::mutate(id = dplyr::row_number(), treatment = regimen)

t_end <- 240 # 10 days: many terminal half-lives even at the lowest CRCL
nca_ev <- dplyr::bind_rows(
  nca_subjects |>
    dplyr::rowwise() |>
    dplyr::reframe(id = id, time = seq(0, t_end - tau, by = tau), amt = dose, rate = 1000, evid = 1L),
  nca_subjects |>
    dplyr::rowwise() |>
    dplyr::reframe(id = id, time = seq(t_end - tau, t_end, by = 0.25), amt = NA_real_, rate = NA_real_, evid = 0L)
) |>
  dplyr::mutate(cmt = "central") |>
  dplyr::left_join(nca_subjects |> dplyr::select(id, WT, CRCL, treatment, tau, dose), by = "id") |>
  dplyr::arrange(id, time, dplyr::desc(evid))
stopifnot(!anyDuplicated(unique(nca_ev[, c("id", "time", "evid")])))

rxode2::rxSetSeed(20211)
sim_nca_raw <- rxode2::rxSolve(
  mod, events = nca_ev, keep = c("treatment", "tau", "dose"),
  rtol = 1e-10, atol = 1e-12
) |>
  as.data.frame()
stopifnot(!anyNA(sim_nca_raw$Cc))

# Re-anchor the last dosing interval at time 0 so PKNCA sees a single dose.
sim_nca <- sim_nca_raw |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(time = time - (t_end - tau)) |>
  dplyr::select(id, time, Cc, treatment, tau, cl)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)
dose_df <- nca_subjects |>
  dplyr::transmute(id, treatment, time = 0, amt = dose, tau)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)
intervals <- tibble::tibble(
  treatment = c("1 g q12h", "1 g q8h"),
  start = 0,
  end = c(12, 8),
  cmax = TRUE,
  cmin = TRUE,
  auclast = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

auc_check <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::left_join(sim_nca |> dplyr::distinct(id, cl), by = "id") |>
  dplyr::mutate(closed_form = 1000 / cl, pct = 100 * (PPORRES / closed_form - 1))
stopifnot(nrow(auc_check) == 2L * n_nca)
# Same drawn parameters on both sides: only trapezoid error separates them.
stopifnot(max(abs(auc_check$pct)) < 1)

reference_auc <- auc_check |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(auclast = median(closed_form), .groups = "drop")

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = reference_auc,
  by = "treatment",
  params = "auclast",
  units = c(auclast = "ug*h/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Steady-state AUC over one dosing interval: PKNCA versus Dose / CL (median).")

as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "cmin")) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(median = median(PPORRES), p05 = quantile(PPORRES, 0.05), p95 = quantile(PPORRES, 0.95), .groups = "drop") |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 1))) |>
  dplyr::rename("Regimen" = treatment, "Parameter" = PPTESTCD, "Median (ug/mL)" = median, "5th pct" = p05, "95th pct" = p95) |>
  knitr::kable(caption = "Steady-state Cmax and Cmin in the model-population virtual cohort.")
```

For context, the observed concentrations in the model population (Table 2)
were troughs of 9.07 +/- 5.47 ug/mL and peaks, drawn 0.5-1 h after the end of
infusion, of 32.03 +/- 10.45 ug/mL, pooled over all regimens and sampling
days. They are not NCA values and are not gated.

# Assumptions and deviations

- **Residual error.** Fu 2021 compared additive, proportional and combined
  residual-error models and carried the combined form into the base model,
  but reports no magnitude for either component in any model. Both `propSd`
  and `addSd` are held at zero; simulations are typical- or individual-level
  predictions without measurement noise. Supply your own values for VPC-style
  use.
- **IIV on Q and V2.** Equations (8)-(9) include eta3 and eta4, but Table 3
  and the base-model report give magnitudes only for eta1 (CL) and eta2 (V1).
  `etalq` and `etalvp` are held at zero rather than invented.
- **IIV scale.** Table 3 lists eta1 = 17.8% and eta2 = 33% without saying
  whether they are sqrt(omega^2) x 100 or a log-normal CV. They are read as
  sqrt(omega^2) x 100 (variance = 0.031684 and 0.1089), consistent with the
  base model's direct omega^2 report; the alternative log(1 + CV^2) reading
  gives 0.0312 and 0.1033, a difference that no reported quantity resolves.
- **Allometric exponents.** 0.75 (CL, Q) and 1 (V1, V2) appear in the printed
  equations with no estimate, RSE or interval and are encoded as fixed.
- **Figure 3 simulation settings.** The paper does not state the weight,
  infusion duration or trough time of its Monte Carlo simulation. The
  maintainers used 70 kg (the reference weight), 1000 mg/h infusions (the
  Table 2 maximum and modal rate) and a trough 30 min before the 48-h dose
  (Results 3.3 and 3.4). This combination reproduces all 16 published medians
  to within about 5%; the cohort-mean weight of 57 kg does not.
- **Figure 3 digitisation.** Box medians were read by the maintainers from the
  figure at 300 dpi, calibrated on the dotted 10/15/20 ug/mL reference lines
  (about 0.1 ug/mL resolution).
- **Table 5 cohorts.** Weight and CRCL were sampled independently from
  truncated normal distributions with the Table 5 means, SDs and ranges; their
  correlation is not reported. The simulated mean troughs fall about 10-15%
  below the observed means in both arms (a known deviation, discussed above).
- **Units.** Doses are in mg and volumes in L, so `Cc` is in mg/L, identical to
  the ug/mL used throughout the paper.
- **Errata.** Europe PMC lists no correction linked to Fu 2021 (PMID
  34651308), and a search for a correction citing its DOI returned nothing, as
  of 2026-10-03.
