---
title: "Dabigatran coagulation-time PD (Fuhr 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dabigatran coagulation-time PD (Fuhr 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Fuhr_2020_dabigatran"))$reference`
* Description: `r rxode2::rxode(readModelDb("Fuhr_2020_dabigatran"))$description`
* Article: <https://doi.org/10.1007/s40262-019-00857-y> (open access, PMC7292816)

Fuhr 2020 describes how dabigatran is reversed by its antidote idarucizumab and
by hemodialysis. The analysis has five parts:

1. a whole-body PBPK model of idarucizumab, including an extended kidney with
   megalin-mediated tubular reabsorption;
2. the idarucizumab-dabigatran binding interaction (Kd 2.1 pmol/L), merged with
   the authors' earlier dabigatran PBPK model;
3. a hemodialysis extension with a dialyzer compartment;
4. a **direct concentration-effect PD model** linking the unbound sum dabigatran
   plasma concentration to four coagulation times (aPTT, dTT, ECT and TT);
5. simulations of reversal scenarios.

Parts 1-3 were built in PK-Sim / MoBi (Open Systems Pharmacology Suite 8.0).
**Only part 4, the PD layer, is packaged in nlmixr2lib.** The reasons are in
[Assumptions and deviations](#assumptions-and-deviations). The PD layer does not
depend on the PBPK model's structure. Its only input is the unbound sum
dabigatran concentration, which you supply as the time-varying covariate
`CU_DABIGATRAN_UM` (umol/L). You can take it from any dabigatran PK model or
from measured concentrations.

# Population

The PD equations were fitted to digitised study-arm means pooled from 26
dabigatran-treated arms of three clinical programmes (Fuhr 2020 ESM Table S8):

* Glund 2015: healthy Caucasian men.
* Glund 2016 / 2017: middle-aged (45-64 years) and elderly (65-80 years)
  Caucasians, and Caucasians with mild or moderate renal impairment.
* Yasaka 2017: healthy Japanese men.

All subjects received steady-state oral dabigatran etexilate: 220 mg twice
daily, or 150 mg twice daily in renal impairment, for 3.5 days. Most arms then
received intravenous idarucizumab (1000-7500 mg), and the fitted data include
coagulation measurements both before and after the antidote. Across the Table 1
arms, mean age was 24-72 years, mean weight 57-90 kg and mean creatinine
clearance 59-139 mL/min. The number of individuals behind the PD fit is not
reported.

# Source trace

| Model element | Value | Source |
|---|---|---|
| Driver `CU_DABIGATRAN_UM` = unbound dabigatran + unbound glucuronide | umol/L | ESM Eq. 3; main text 2.3 |
| aPTT = Emax * C / (EC50 + C) + slope * C + baseline | | ESM Eq. 16; main text 3.4 |
| `lrbase_aPTT` | 30.21 s | ESM Eq. 16 |
| `lemax_aPTT` | 17.94 s | ESM Eq. 16 |
| `lec50_aPTT` | 0.04 umol/L | ESM Eq. 16 |
| `lslope_aPTT` | 61.67 s per umol/L | ESM Eq. 16 |
| dTT = slope * C + baseline | | ESM Eq. 17 |
| `lrbase_dTT` | 31.59 s | ESM Eq. 17 |
| `lslope_dTT` | 88.30 s per umol/L | ESM Eq. 17 |
| ECT = slope * C + baseline | | ESM Eq. 18 |
| `lrbase_ECT` | 34.85 s | ESM Eq. 18 |
| `lslope_ECT` | 209.70 s per umol/L | ESM Eq. 18 |
| TT = Emax * C / (EC50 + C) + slope * C + baseline | | ESM Eq. 19; main text 3.4 |
| `lrbase_TT` | 12.93 s | ESM Eq. 19 |
| `lemax_TT` | 106.28 s | ESM Eq. 19 |
| `lec50_TT` | 0.19 umol/L | ESM Eq. 19 |
| `lslope_TT` | 144.02 s per umol/L | ESM Eq. 19 |
| `addSd_aPTT`, `addSd_dTT`, `addSd_ECT`, `addSd_TT` | 0 (fixed) | not reported |

# Concentration-effect curves

```{r sim}
mod <- rxode2::rxode2(readModelDb("Fuhr_2020_dabigatran"))

mw_dab <- 471.5 # g/mol; the ESM rounds it to 472

conc_ngml <- seq(0, 220, by = 2)
conc_grid <- data.frame(
  id = 1L,
  time = seq_along(conc_ngml) - 1,
  evid = 0L,
  amt = 0,
  dvid = 1L,
  conc_ngml = conc_ngml
) |>
  mutate(CU_DABIGATRAN_UM = conc_ngml / mw_dab)

sim <- rxode2::rxSolve(
  mod,
  events = conc_grid,
  omega = NA,
  sigma = NA,
  keep = "conc_ngml",
  returnType = "data.frame"
)

sim_long <- sim |>
  dplyr::select(conc_ngml, aPTT, dTT, ECT, TT) |>
  tidyr::pivot_longer(-conc_ngml, names_to = "assay", values_to = "seconds") |>
  mutate(assay = factor(assay, levels = c("aPTT", "dTT", "ECT", "TT")))
```

```{r fig6, fig.width = 7, fig.height = 5}
# Replicates Figure 6 (left column) of Fuhr 2020: coagulation times versus the
# unbound sum dabigatran plasma concentration.
ggplot(sim_long, aes(conc_ngml, seconds, colour = assay)) +
  geom_line(linewidth = 1) +
  facet_wrap(~assay, scales = "free_y") +
  labs(
    x = "Unbound sum dabigatran plasma concentration (ng/mL)",
    y = "Coagulation time (s)",
    caption = "Replicates Figure 6 (left column) of Fuhr 2020."
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

# Validation

The PD layer is a set of closed-form equations with no dosing, so NCA does not
apply. The model is checked in three ways instead.

## Baselines at zero concentration

With no dabigatran, every assay must return its intercept exactly. This checks
the transcription of the four baselines.

```{r baseline}
at0 <- sim |> dplyr::filter(conc_ngml == 0)
stopifnot(
  isTRUE(all.equal(at0$aPTT, 30.21)),
  isTRUE(all.equal(at0$dTT, 31.59)),
  isTRUE(all.equal(at0$ECT, 34.85)),
  isTRUE(all.equal(at0$TT, 12.93))
)
at0 |> dplyr::select(aPTT, dTT, ECT, TT) |> knitr::kable(digits = 2)
```

## Agreement with the published fitted curves (unit check)

The ESM gives the concentration unit only in its definition of the driver
(Equation 3, umol/L), while Figure 6 plots the curves against ng/mL. The
maintainers read the published fitted lines off Figure 6 at 100 and 220 ng/mL,
to about +/- 2 s. If the equations had been in ng/mL or nmol/L, the predictions
would be off by orders of magnitude, so agreement here confirms the umol/L
reading and the dabigatran molecular-weight conversion.

```{r fig6-check}
fig6 <- data.frame(
  conc_ngml = rep(c(100, 220), each = 4),
  assay = rep(c("aPTT", "dTT", "ECT", "TT"), 2),
  figure6_s = c(59, 50, 80, 98, 75, 72, 132, 155)
)

chk <- fig6 |>
  left_join(sim_long |> mutate(assay = as.character(assay)), by = c("conc_ngml", "assay")) |>
  mutate(pct_diff = 100 * (seconds - figure6_s) / figure6_s)

chk |>
  dplyr::rename(
    "Unbound sum DAB (ng/mL)" = conc_ngml,
    "Assay" = assay,
    "Figure 6 curve (s)" = figure6_s,
    "Model (s)" = seconds,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = 1)

# Deterministic comparison of a closed-form curve with its own published plot;
# the only error is the maintainers' reading of the figure (about 2 s).
stopifnot(all(abs(chk$pct_diff) < 5))
```

## Cross-assay consistency in the individual profile

The middle column of Figure 6 shows the four model-predicted coagulation times
for healthy Japanese subjects on 220 mg dabigatran etexilate twice daily, before
and after 1000 mg idarucizumab at 242 h. All four curves come from one predicted
concentration, so the peak just before idarucizumab must map to a single
concentration under all four equations. The maintainers read the ECT peak as
about 115 s. Inverting Equation 18 gives the concentration, from which the other
three assays follow. The Figure 6 peaks for comparison are about 70 s (aPTT),
65 s (dTT) and 145 s (TT).

```{r consistency}
c_peak <- (115 - 34.85) / 209.70 # umol/L, from the ECT peak
peak <- rxode2::rxSolve(
  mod,
  events = data.frame(id = 1L, time = 0, evid = 0L, amt = 0, dvid = 1L, CU_DABIGATRAN_UM = c_peak),
  omega = NA,
  sigma = NA,
  returnType = "data.frame"
)
peak_chk <- data.frame(
  assay = c("aPTT", "dTT", "TT"),
  figure6_s = c(70, 65, 145),
  model_s = c(peak$aPTT, peak$dTT, peak$TT)
) |>
  mutate(pct_diff = 100 * (model_s - figure6_s) / figure6_s)
knitr::kable(peak_chk, digits = 1)
c_peak * mw_dab # ng/mL
stopifnot(all(abs(peak_chk$pct_diff) < 10))
```

After idarucizumab, Figure 6 shows every assay falling to its baseline (about
31, 31, 35 and 13 s). This matches the zero-concentration values above, which
is expected because idarucizumab captures essentially all unbound dabigatran.

# Illustrative time course

The model needs a concentration trajectory. As a demonstration only, this
section takes the typical-value total dabigatran profile from the Yang 2024
popPK model in nlmixr2lib (healthy Chinese adults, fasted, ABCB1 rs4148738
non-CT). It runs 220 mg twice daily for 3.5 days (seven doses, as in the Fuhr
2020 studies) and converts the result to unbound concentrations with fu = 0.65
(35% protein binding, Fuhr 2020 Introduction). This is not the Fuhr 2020 PBPK
prediction. Yang 2024 was fitted to single 150 mg doses, and its total
dabigatran may not match the hydrolysed "sum" moiety.

```{r timecourse, fig.width = 7, fig.height = 5}
pk <- rxode2::rxode2(readModelDb("Yang_2024_dabigatran"))
pk_ev <- rxode2::et(amt = 220, cmt = "depot", ii = 12, addl = 6) |>
  rxode2::et(seq(0, 96, by = 0.25), cmt = "central") |>
  as.data.frame() |>
  mutate(FED_HIGHFAT = 0, SNP_ABCB1_RS4148738_HET = 0)
pk_sim <- rxode2::rxSolve(pk, events = pk_ev, omega = NA, sigma = NA, returnType = "data.frame")

pd_ev <- data.frame(
  id = 1L,
  time = pk_sim$time,
  evid = 0L,
  amt = 0,
  dvid = 1L,
  CU_DABIGATRAN_UM = 0.65 * pk_sim$Cc / mw_dab
)
pd_sim <- rxode2::rxSolve(mod, events = pd_ev, omega = NA, sigma = NA, returnType = "data.frame")

pd_sim |>
  dplyr::select(time, aPTT, dTT, ECT, TT) |>
  tidyr::pivot_longer(-time, names_to = "assay", values_to = "seconds") |>
  ggplot(aes(time, seconds, colour = assay)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~assay, scales = "free_y") +
  labs(
    x = "Time (h)",
    y = "Coagulation time (s)",
    caption = "Illustrative only: Yang 2024 typical PK, fu = 0.65, driving Fuhr 2020 PD."
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

# Assumptions and deviations {#assumptions-and-deviations}

## The PBPK, binding and dialysis layers are not packaged

The idarucizumab and dabigatran PBPK models run on the PK-Sim / MoBi platform.
Their organ volumes, blood flows, surface areas and tissue partition
coefficients come from the platform's physiology database for each simulated
individual and are not printed in the paper or the ESM. The ESM prints the
drug-specific parameters (Table 2 and Table S3), the kidney-extension equations
(ESM Eq. 5-13), the binding kinetics (ESM Eq. 14-15) and the dialyzer clearance
(main text Eq. 2, with KoA from Liesenfeld 2013). The whole-body structure they
plug into cannot be rebuilt from the publication, so these layers are not
reproduced. The binding and dialysis parts act on platform-computed plasma and
interstitial concentrations, so they cannot be separated from the PBPK model
either.

A dabigatran hemodialysis model with the same dialysis-clearance form is
available as `Liesenfeld_2013_dabigatran`.

## Driver definition

`CU_DABIGATRAN_UM` is the unbound sum of dabigatran and its acyl-glucuronide.
It excludes drug bound to plasma proteins and drug bound to idarucizumab. This
matches the assays in the source studies, which hydrolysed the glucuronide and
ultrafiltered the sample. A total plasma concentration must be multiplied by
the unbound fraction (about 0.65) before use. A value in ng/mL must be divided
by 471.5.

## No IIV and no residual error

The paper fitted the equations to digitised study-arm means and reports only
point estimates, with no standard errors, inter-individual variability or
residual-error model. Performance is summarised as mean relative deviation
(1.16, 1.11, 1.16 and 1.27 for aPTT, dTT, ECT and TT). All four additive
residual SDs are therefore fixed at zero. Free them when refitting to
individual data.

## Figure readings

The Figure 6 values used in the validation were read by the maintainers from
the published figure. They serve as unit and transcription checks only; no
model value was taken from them.
