---
title: "Soybean oil-based lipid emulsion (Fukushima 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Soybean oil-based lipid emulsion (Fukushima 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Fukushima_2022_soybeanOilLipidEmulsion"))$reference`
* Description: `r rxode2::rxode(readModelDb("Fukushima_2022_soybeanOilLipidEmulsion"))$description`
* Article (open access): <https://doi.org/10.1002/jpen.2111>
  (PMC9290941); the Supplementary Materials document is distributed with the
  article on Europe PMC.

Fukushima et al. describe plasma triglyceride (TG) during and after an
infusion of a 20% soybean oil-based intravenous lipid emulsion (SO-ILE,
Intralipos 20%) with a single plasma compartment (Eq 1):

$$\frac{dA}{dt} = \mathrm{EndVLDL} - A \cdot K_{el} + \mathrm{InfRate}, \qquad C = A / V$$

where `EndVLDL` is the zero-order endogenous TG (VLDL) production rate,
`Kel` the first-order elimination rate constant, `InfRate` the emulsion
infusion rate and `V` the plasma volume, fixed at 0.422 dL/kg. In the model
file `EndVLDL` is `kin`, `Kel` is `kel` and `V` is `vc`; the infusion is an
ordinary zero-order infusion (`amt` and `rate`) into `central`, with `amt` in
mg of TG (1 g of fat = 1000 mg of TG). The paper uses the model to derive an
individual maximum acceptable infusion rate (Eq 7) and a nomogram (Figure 2).

# Population

Eighty-three adult Japanese inpatients (44 male / 39 female) receiving
peripheral (n = 75) or central venous (n = 8) parenteral nutrition at two
hospitals in Japan (October 2016 - March 2018) were analysed (Table 1). The
median age was 74 years (range 20-95) and median body weight 55.8 kg
(31.2-86.1). Most had acute inflammatory gastrointestinal disease (colon
diverticulitis, acute cholecystitis, peptic ulcer, ischemic colitis,
colorectal cancer, ...) with low serum albumin (median 3.52 g/dL) and raised
CRP (median 1.56 mg/dL). Baseline TG had a median of 94 mg/dL (30-291);
patients with baseline TG above 300 mg/dL were excluded. Each patient
received one SO-ILE infusion of 20-120 g fat (median 50 g) at 0.053-0.363
g/kg/h (median 0.217) for 0.75-19.8 h (median 5.2 h). Two TG samples were
drawn during the infusion (the first within 0-1, 1-2 or 2-3 h by randomised
group, the second at the end of infusion), and 72 next-morning routine
samples were added, giving 238 concentrations.

The same information is available programmatically via
`readModelDb("Fukushima_2022_soybeanOilLipidEmulsion")()$population`.

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `d/dt(central) <- kin - kel * central` (+ infusion) | n/a | Eq 1; Supplementary Scheme 2, Model 3 |
| `Cc <- central / vc` | n/a | Methods, 'C = A / V' |
| `central(0) <- TRIG * vc` | n/a | Not printed; established from Supplementary Figure S7 (see below) |
| `vc = 0.422 * WT` (`lvc`, fixed) | 0.422 dL/kg | Table 4 final model structure; Methods (fixed from healthy-subject literature) |
| `lkin` (EndVLDL, theta1) | 5709 mg/h | Table 4 |
| `lkel` (Kel, theta2) | 2.51 1/h | Table 4 |
| `e_trig_kel` (theta3) | -0.833 | Table 4; reference TG 94 mg/dL (Table 4 note) |
| `e_wt_kel` (theta4) | -1.27 | Table 4; reference WT 55.8 kg (Table 4 note) |
| `etalkel` | 0.433^2 = 0.1875 | Table 4 'omega Kel 43.3%', Eq 2 (exponential IIV) |
| `propSd` | 0.318 | Table 4 'delta 31.8%', Eq 3 'CObs = C x (1 + eps)' |
| No IIV on EndVLDL or V | n/a | Table 3 model 9; Supplementary 'Model Selection' |
| Eq 6, `Cmax = (EndVLDL + InfRate) / (V x Kel)` | n/a | Methods, Eq 6 |
| Eq 7, maximum acceptable infusion rate | n/a | Results, Eq 7 (target mean Cmax 211 mg/dL = 400 / 1.894) |

# Typical-patient profile and the initial condition (Supplementary Figure S7)

The paper does not print the initial condition of Eq 1. Two choices are
natural: start at the measured baseline, `A(0) = TGbase x V`, or start at
the model's own steady state, `A(0) = EndVLDL / Kel`. For the typical
patient they coincide to within 3% (94 vs 96.6 mg/dL), but they spread
differently across patients: the steady-state start inherits the full
between-patient variability of `Kel`, the baseline start does not.

Supplementary Figure S7 shows the median and 90th percentile of 1,000
simulated profiles for a typical patient (TGbase 94 mg/dL, 55.8 kg) given
0.121 g/kg/h for 6 h, with the residual error included. At the first plotted
time (about 0.1 h), the 90th-percentile curve for V = 0.422 x BW starts at
about 165 mg/dL (all five plasma-volume variants start between about 155 and
185 mg/dL). The percentiles below are computed without Monte Carlo noise.
The typical-value model is solved on a deterministic grid of 200 quantiles of
`etalkel`, and the proportional residual error is then integrated out
analytically.

```{r s7-grid}
mod <- readModelDb("Fukushima_2022_soybeanOilLipidEmulsion")
mod_ui <- rxode2::rxode2(mod)
mod_typ <- rxode2::zeroRe(mod_ui)

ini_df <- mod_ui$iniDf
omega_kel <- sqrt(ini_df$est[ini_df$name == "etalkel"])
prop_sd <- ini_df$est[ini_df$name == "propSd"]
lkel_typ <- ini_df$est[ini_df$name == "lkel"]

# Deterministic quadrature grid over the between-patient variability of kel
eta_grid <- omega_kel * qnorm(ppoints(200))

wt_typ <- 55.8
tg_typ <- 94
rate_s7 <- 0.121 * wt_typ * 1000 # mg TG / h
t_obs <- c(0, 0.1, 0.25, 0.5, 1, 2, 3, 4, 5, 6, 6.25, 6.5, 7, 7.5, 8, 9)

make_grid_events <- function(rate_mg_h, dur_h, times, tg, wt, n) {
  dose <- data.frame(
    id = seq_len(n), time = 0, evid = 1L, amt = rate_mg_h * dur_h,
    rate = rate_mg_h, cmt = "central"
  )
  obs <- expand.grid(id = seq_len(n), time = times) |>
    dplyr::mutate(evid = 0L, amt = 0, rate = 0, cmt = "central")
  dplyr::bind_rows(dose, obs) |>
    dplyr::mutate(TRIG = tg, WT = wt) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

ev_grid <- make_grid_events(rate_s7, 6, t_obs, tg_typ, wt_typ, length(eta_grid))
par_grid <- data.frame(lkel = lkel_typ + eta_grid)

# The grid solves are deliberately multi-subject with zeroed omega (each
# subject is one quadrature node); muffle only that expected warning.
solve_quadrature <- function(model, events, params) {
  withCallingHandlers(
    rxode2::rxSolve(
      model,
      events = events, params = params,
      rtol = 1e-10, atol = 1e-12, returnType = "data.frame"
    ),
    warning = function(w) {
      if (grepl("without 'omega'", conditionMessage(w), fixed = TRUE)) {
        invokeRestart("muffleWarning")
      }
    }
  )
}
solve_grid <- function(model) solve_quadrature(model, ev_grid, par_grid)

# Quantile of C x (1 + eps), eps ~ N(0, propSd^2), mixed over the eta grid
mix_quantile <- function(conc, p, sd = prop_sd) {
  stats::uniroot(
    function(x) mean(stats::pnorm((x / conc - 1) / sd)) - p,
    interval = c(0, 10 * max(conc))
  )$root
}

summarise_grid <- function(sim) {
  sim |>
    dplyr::group_by(time) |>
    dplyr::summarise(
      median = mix_quantile(Cc, 0.5),
      p90 = mix_quantile(Cc, 0.9),
      .groups = "drop"
    )
}

# Shipped model: central(0) <- TRIG * vc
s7_base <- summarise_grid(solve_grid(mod_typ)) |>
  dplyr::mutate(start = "A(0) = TGbase x V (shipped)")

# Alternative reading: start at the model steady state
mod_ss <- mod_typ |> rxode2::model(central(0) <- kin / kel)
s7_ss <- summarise_grid(solve_grid(mod_ss)) |>
  dplyr::mutate(start = "A(0) = EndVLDL / Kel (alternative)")

s7 <- dplyr::bind_rows(s7_base, s7_ss)
s7 |>
  dplyr::filter(time %in% c(0.1, 1, 6, 9)) |>
  dplyr::arrange(start, time) |>
  dplyr::rename(
    "Initial condition" = start,
    "Time (h)" = time,
    "Median (mg/dL)" = median,
    "90th percentile (mg/dL)" = p90
  ) |>
  knitr::kable(digits = 1)
```

```{r s7-figure, fig.width = 7, fig.height = 4.5}
s7 |>
  tidyr::pivot_longer(c(median, p90), names_to = "stat", values_to = "Cc") |>
  ggplot(aes(time, Cc, colour = start, linetype = stat)) +
  geom_line() +
  annotate("point", x = 0.1, y = 165, shape = 4, size = 3) +
  labs(
    x = "Time after start of infusion (h)", y = "TG concentration (mg/dL)",
    colour = NULL, linetype = NULL,
    title = "Typical patient, 0.121 g/kg/h for 6 h",
    caption = paste(
      "Replicates Supplementary Figure S7 of Fukushima 2022 (V = 0.422 x BW).",
      "Cross: 90th percentile at 0.1 h read from the figure."
    )
  ) +
  theme(legend.position = "bottom", legend.direction = "vertical")
```

```{r s7-checks}
p90_early <- function(x) x$p90[x$time == 0.1]
med_at <- function(x, t) x$median[x$time == t]
stopifnot(
  # Early 90th percentile: the shipped start matches the figure (~165 mg/dL);
  # the steady-state start (about 217 mg/dL) cannot.
  abs(p90_early(s7_base) / 165 - 1) < 0.10,
  abs(p90_early(s7_ss) / 165 - 1) > 0.20,
  # Plateau 90th percentile ~400 mg/dL, the design target of the dosing rule.
  abs(s7_base$p90[s7_base$time == 6] / 400 - 1) < 0.02,
  # Medians read from Figure S7: ~210 mg/dL plateau, ~97 mg/dL at 9 h. The
  # exact medians here are 204 and 94 (proportional error on a log-normal
  # kel pulls the median below the typical value); the figure's 1,000-subject
  # Monte Carlo median carries an SE of ~2% (~4.6 mg/dL at the plateau).
  abs(med_at(s7_base, 6) / 210 - 1) < 0.06,
  abs(med_at(s7_base, 9) / 97 - 1) < 0.06
)
```

The baseline start reproduces the early spread in Figure S7, and the
steady-state start overshoots it by about a third. The model therefore
starts each patient at the measured baseline TG. After the infusion stops,
TG returns to the model steady state `EndVLDL / (Kel x V)`, not to `TGbase`.
For the typical patient that is 96.6 mg/dL, consistent with the ~97 mg/dL
tail of Figure S7. The plateau of the typical-value profile is Eq 6,
210.8 mg/dL (checked to 0.2% in the PKNCA section below). The median in the
table is lower (204 mg/dL) because a proportional residual error applied to
a log-normal `Kel` pulls the median below the typical value. Figure S7's
plateau median (~210 mg/dL) comes from 1,000 simulated patients and lies
within its Monte Carlo noise.

# Variability propagation: the 1.894 ratio and the omega scale

The paper reports that, by 10,000-patient Monte Carlo simulation, the 90th
percentile of Cmax is 1.894 +/- 0.014 times the population-mean Cmax of
Eq 6, so the target mean Cmax is 400 / 1.894 = 211 mg/dL. This ratio depends
only on the between-patient and residual variability. Table 4 gives
"omega Kel = 43.3%", which could be the SD of `eta` (variance 0.1875) or a
CV% (variance log(1 + 0.433^2) = 0.172). The ratio is computed on the same
quadrature grid at the end of a 6-h infusion, when TG is at steady state for
every patient (the slowest grid patient has a half-life of about 1 h):

```{r ratio}
ratio_for_omega <- function(omega) {
  ev <- make_grid_events(rate_s7, 6, 6, tg_typ, wt_typ, 200)
  sim <- solve_quadrature(
    mod_typ, ev,
    data.frame(lkel = lkel_typ + omega * qnorm(ppoints(200)))
  )
  typical <- (exp(ini_df$est[ini_df$name == "lkin"]) + rate_s7) /
    (exp(lkel_typ) * 0.422 * wt_typ) # Eq 6
  mix_quantile(sim$Cc, 0.9) / typical
}
ratio_sd <- ratio_for_omega(omega_kel)
ratio_cv <- ratio_for_omega(sqrt(log(1 + 0.433^2)))
data.frame(
  reading = c("omega = 0.433 is the SD of eta (shipped)", "43.3% is a CV"),
  ratio = c(ratio_sd, ratio_cv),
  published = 1.894
) |>
  dplyr::rename(
    "Reading of 'omega Kel 43.3%'" = reading,
    "p90 / mean Cmax" = ratio,
    "Published" = published
  ) |>
  knitr::kable(digits = 3)

stopifnot(
  # The SD reading lands within the paper's +/-0.014 Monte Carlo spread;
  # the CV reading is two spreads away.
  abs(ratio_sd - 1.894) < 0.014,
  abs(ratio_cv - 1.894) > 0.025
)
```

# Maximum acceptable infusion rate (Eq 7 and Supplementary Table S6)

Eq 7 rearranges Eq 6 for the infusion rate that gives a mean Cmax of
211 mg/dL. Here it is evaluated with the packaged model's individual `kel`
and `vc`, read from a typical-value solve:

```{r eq7}
mod_params <- function(tg, wt) {
  ev <- data.frame(
    id = seq_along(tg), time = 0, evid = 0L, amt = 0, cmt = "central",
    TRIG = tg, WT = wt
  )
  solve_quadrature(mod_typ, ev, NULL) |>
    dplyr::select(kin, kel, vc)
}
# Eq 7, expressed per kg (g/kg/h)
max_rate <- function(tg, wt, target = 211) {
  p <- mod_params(tg, wt)
  (target * p$kel * p$vc - p$kin) / 1000 / wt
}
rate_typ <- max_rate(tg_typ, wt_typ)
rate_typ
```

Supplementary Table S6 refits the model with V fixed at 0.3-0.6 dL/kg and
reports the typical-patient Eq 7 rate for each fit. The 0.422 dL/kg column is
the packaged model. The other columns check that Eq 7 has been read
correctly (same formula, Table S6 estimates):

```{r tableS6}
s6 <- tibble::tribble(
  ~v_per_kg, ~theta1, ~theta2, ~theta3, ~theta4, ~published,
  0.3,       6005,    3.70,    -0.826,  -1.262,  0.127,
  0.4,       5770,    2.67,    -0.833,  -1.270,  0.122,
  0.5,       5540,    2.06,    -0.835,  -1.273,  0.118,
  0.6,       5336,    1.65,    -0.840,  -1.286,  0.113
) |>
  dplyr::mutate(
    eq7 = (211 * theta2 * (tg_typ / 94)^theta3 * (wt_typ / 55.8)^theta4 *
      v_per_kg * wt_typ - theta1) / 1000 / wt_typ
  ) |>
  dplyr::bind_rows(
    tibble::tibble(v_per_kg = 0.422, published = 0.121, eq7 = rate_typ)
  ) |>
  dplyr::arrange(v_per_kg)

s6 |>
  dplyr::select(v_per_kg, eq7, published) |>
  dplyr::rename(
    "V (dL/kg)" = v_per_kg,
    "Eq 7 rate (g/kg/h)" = eq7,
    "Table S6 (g/kg/h)" = published
  ) |>
  knitr::kable(digits = 4)

# Table S6 prints three decimals, so a correct reading is within 0.0005.
stopifnot(all(abs(s6$eq7 - s6$published) < 0.0006))
```

# Nomogram (Figure 2)

Figure 2 draws the Eq 7 rate as contours over baseline TG and body weight.
The contour heights below were digitised by the maintainers from the
published figure at three body weights:

```{r figure-2}
digitised <- tibble::tribble(
  ~wt,  ~rate, ~tg,
  30.6, 0.300,  97.4, 30.6, 0.200, 128.5, 30.6, 0.150, 151.8,
  30.6, 0.125, 166.5, 30.6, 0.100, 184.5, 30.6, 0.075, 205.9,
  30.6, 0.050, 232.6, 30.6, 0.025, 266.0, 30.6, 0.010, 290.7,
  40.1, 0.300,  72.1, 40.1, 0.200,  98.1, 40.1, 0.150, 118.8,
  40.1, 0.125, 132.5, 40.1, 0.100, 149.1, 40.1, 0.075, 169.8,
  40.1, 0.050, 197.2, 40.1, 0.025, 232.9, 40.1, 0.010, 261.0,
  89.4, 0.200,  39.0, 89.4, 0.150,  50.4, 89.4, 0.125,  58.7,
  89.4, 0.100,  69.7, 89.4, 0.075,  85.4, 89.4, 0.050, 108.4,
  89.4, 0.025, 146.1, 89.4, 0.010, 182.9
)

# Baseline TG at which the packaged model's Eq 7 rate equals each contour
tg_on_contour <- function(rate, wt) {
  stats::uniroot(
    function(tg) max_rate(tg, wt) - rate,
    interval = c(1, 3000), tol = 1e-6
  )$root
}
digitised$tg_model <- mapply(tg_on_contour, digitised$rate, digitised$wt)
digitised$tg_ratio <- digitised$tg / digitised$tg_model

grid <- expand.grid(wt = seq(30, 90, by = 1), tg = seq(30, 300, by = 2))
grid$rate <- max_rate(grid$tg, grid$wt)

ggplot(grid, aes(wt, tg)) +
  geom_contour(
    aes(z = rate),
    breaks = c(0.01, 0.025, 0.05, 0.075, 0.1, 0.125, 0.15, 0.2, 0.3),
    colour = "black"
  ) +
  geom_point(data = digitised, colour = "red", shape = 4) +
  annotate("point", x = wt_typ, y = tg_typ, shape = 16, size = 2.5, colour = "blue") +
  coord_cartesian(xlim = c(30, 90), ylim = c(30, 300)) +
  labs(
    x = "Body weight (kg)", y = "Baseline TG (mg/dL)",
    title = "Eq 7 contours, 0.01-0.3 g/kg/h",
    caption = paste(
      "Replicates Figure 2 of Fukushima 2022. Lines: packaged model;",
      "red crosses: digitised contours; blue dot: typical patient."
    )
  )

knitr::kable(
  digitised |>
    dplyr::group_by(wt) |>
    dplyr::summarise(
      min_ratio = min(tg_ratio), median_ratio = median(tg_ratio),
      max_ratio = max(tg_ratio), .groups = "drop"
    ) |>
    dplyr::rename(
      "Body weight (kg)" = wt,
      "Min ratio" = min_ratio,
      "Median ratio" = median_ratio,
      "Max ratio" = max_ratio
    ),
  digits = 3,
  caption = "Digitised / model baseline TG on each contour, by body weight."
)

stopifnot(
  # Shape: one constant ratio across all 26 digitised contour points means
  # the TG exponent (theta3) and body-weight exponent (theta4) carry through
  # unchanged.
  stats::sd(log(digitised$tg_ratio)) < 0.02,
  # Offset: the published contours sit a uniform ~5% higher in baseline TG
  # (see Assumptions and deviations).
  abs(median(digitised$tg_ratio) - 1.055) < 0.02
)
```

The digitised contours lie on the model's contours up to one uniform factor
of about 1.055 in baseline TG. A single constant factor across rates and
weights is what a slightly different target Cmax or `Kel` in the drawing
would produce: `(1.055)^0.833` corresponds to about 220 instead of 211 mg/dL.
A misread exponent or a different `EndVLDL` would distort the contour shape
instead. The paper's own tabulated typical-patient rate (0.121 g/kg/h,
Table S6) agrees with Eq 7 and the packaged model, not with the drawn
contours, and the shift is recorded below as a deviation of the published
figure. Because of it, the typical patient (blue dot) falls just above the
0.125 contour here (0.121 g/kg/h). The paper's caption places it in the
"0.125 g/kg/h" area.

# Virtual cohort: observed profiles (Figure 1) and individual rates (Figure 3)

Individual data are not available. The virtual cohort draws each covariate
and dosing variable independently from a piecewise-linear interpolation of
the Table 1 quantiles (minimum, quartiles, maximum).

```{r cohort}
set.seed(2111)
n_sub <- 200L
r_table1 <- function(n, q) {
  stats::approx(c(0, 0.25, 0.5, 0.75, 1), q, xout = stats::runif(n))$y
}
cohort <- tibble::tibble(
  id = seq_len(n_sub),
  WT = r_table1(n_sub, c(31.2, 46.2, 55.8, 65, 86.1)),
  TRIG = r_table1(n_sub, c(30, 68, 94, 119, 291)),
  rate_gkgh = r_table1(n_sub, c(0.053, 0.105, 0.217, 0.278, 0.363)),
  dur = r_table1(n_sub, c(0.75, 3.0, 5.2, 7.2, 19.8))
) |>
  dplyr::mutate(rate = rate_gkgh * WT * 1000, amt = rate * dur)

ev_cohort <- dplyr::bind_rows(
  cohort |>
    dplyr::transmute(id, time = 0, evid = 1L, amt, rate, cmt = "central", TRIG, WT),
  cohort |>
    dplyr::select(id, TRIG, WT, dur) |>
    tidyr::crossing(time = seq(0, 24, by = 0.25)) |>
    dplyr::bind_rows(cohort |> dplyr::transmute(id, TRIG, WT, dur, time = dur)) |>
    dplyr::distinct(id, time, .keep_all = TRUE) |>
    dplyr::transmute(id, time, evid = 0L, amt = 0, rate = 0, cmt = "central", TRIG, WT)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

rxode2::rxSetSeed(2111)
sim_cohort <- rxode2::rxSolve(mod, events = ev_cohort, returnType = "data.frame") |>
  dplyr::left_join(cohort |> dplyr::select(id, dur), by = "id") |>
  dplyr::mutate(phase = ifelse(time <= dur, "during infusion", "after infusion"))
```

```{r figure-1, fig.width = 7, fig.height = 4.5}
ggplot(sim_cohort, aes(time, Cc, group = interaction(id, phase))) +
  geom_line(aes(linetype = phase), alpha = 0.3) +
  scale_linetype_manual(values = c("during infusion" = "solid", "after infusion" = "dashed")) +
  labs(
    x = "Time (h)", y = "Triglyceride concentration (mg/dL)", linetype = NULL,
    title = "Simulated TG profiles, Table 1-like virtual cohort",
    caption = "Compare with Figure 1 of Fukushima 2022 (observed data). Individual predictions without residual error."
  ) +
  theme(legend.position = "bottom")
```

As in Figure 1, most simulated profiles stay below about 500 mg/dL and TG
falls back near baseline within a few hours of the end of the infusion. The
upper tail is thinner than observed. When the maintainers ran this, the
simulated peaks had a median of about 250 mg/dL and only about 1-2% of
patients exceeded 1,000 mg/dL, reaching about 1,500 mg/dL. Figure 1 shows
roughly six of 83 patients above 1,000 mg/dL, reaching about 2,450 mg/dL.
This matches the authors' own finding that the first-order model
underpredicts concentrations above about 1,000 mg/dL, where TG elimination
appears saturated (Supplementary Model Selection and Figure S2).

Figure 3 histograms the Eq 7 rate for the 83 study patients. The same
calculation on the virtual cohort is compared below. The virtual cohort draws
baseline TG and body weight independently, which the real cohort need not
do, so this comparison is descriptive and has no gate:

```{r figure-3}
bins <- c(-Inf, 0.01, 0.025, 0.05, 0.075, 0.1, 0.125, 0.15, 0.2, 0.3, Inf)
bin_labels <- c(
  "<0.01", "0.01-0.025", "0.026-0.05", "0.06-0.075", "0.076-0.1",
  "0.11-0.125", "0.126-0.15", "0.16-0.2", "0.21-0.3", ">=0.3"
)
published_fig3 <- c(3, 2, 3, 10, 11, 7, 10, 7, 15, 15) # Figure 3, n = 83
cohort$max_rate <- max_rate(cohort$TRIG, cohort$WT)
virtual_counts <- as.vector(table(cut(cohort$max_rate, bins, right = FALSE)))
tibble::tibble(
  bin = bin_labels,
  published_pct = 100 * published_fig3 / sum(published_fig3),
  virtual_pct = 100 * virtual_counts / n_sub
) |>
  dplyr::rename(
    "Max acceptable rate (g/kg/h)" = bin,
    "Figure 3, % of 83" = published_pct,
    "Virtual cohort, %" = virtual_pct
  ) |>
  knitr::kable(digits = 1)
```

The virtual cohort over-fills the lowest bins. The interpolated Table 1
quantiles put a quarter of the virtual patients uniformly between 119 and
291 mg/dL baseline TG, which over-represents very high baselines (where Eq 7
falls below 0.01 g/kg/h) compared with the right-skewed real distribution.
Above about 0.05 g/kg/h, the virtual and published distributions are
broadly similar.

# PKNCA validation

The paper reports no NCA. Its one published concentration is the
population-mean Cmax of 211 mg/dL at the typical patient's maximum
acceptable rate. Eq 6 gives the population-mean Cmax at any other infusion
rate. Typical-value profiles are simulated for five 6-hour infusion rates,
and PKNCA Cmax is compared with Eq 6 (and with 211 mg/dL for the 0.121
g/kg/h arm). An infusion of 6 h is about 20 half-lives of the typical `kel`,
so each Cmax is the Eq 6 plateau.

```{r pknca}
arms <- tibble::tibble(rate_gkgh = c(0.05, 0.1, 0.121, 0.2, 0.3)) |>
  dplyr::mutate(
    id = dplyr::row_number(),
    treatment = sprintf("%.3f g/kg/h", rate_gkgh),
    rate = rate_gkgh * wt_typ * 1000,
    amt = rate * 6
  )
ev_nca <- dplyr::bind_rows(
  arms |> dplyr::transmute(id, treatment, time = 0, evid = 1L, amt, rate, cmt = "central"),
  arms |>
    dplyr::select(id, treatment) |>
    tidyr::crossing(time = seq(0, 24, by = 0.25)) |>
    dplyr::mutate(evid = 0L, amt = 0, rate = 0, cmt = "central")
) |>
  dplyr::mutate(TRIG = tg_typ, WT = wt_typ) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim_nca <- rxode2::rxSolve(
  mod_typ,
  events = ev_nca, keep = "treatment",
  rtol = 1e-10, atol = 1e-12, returnType = "data.frame"
) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(
  arms |> dplyr::mutate(time = 0, dur = 6) |>
    dplyr::select(id, treatment, time, amt, dur),
  amt ~ time | treatment + id,
  route = "intravascular", duration = "dur"
)
intervals <- data.frame(start = 0, end = 24, cmax = TRUE, tmax = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

typ <- mod_params(tg_typ, wt_typ)
reference <- arms |>
  dplyr::transmute(
    treatment,
    cmax = ifelse(rate_gkgh == 0.121, 211, (typ$kin + rate) / (typ$kel * typ$vc))
  )

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = reference,
  by = "treatment",
  params = "cmax",
  units = c(cmax = "mg/dL"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = paste(
    "Simulated vs. reference Cmax for a typical patient (94 mg/dL, 55.8 kg).",
    "Reference: 211 mg/dL published for 0.121 g/kg/h; Eq 6 otherwise.",
    "* differs from reference by >20%."
  )
)

cmax_sim <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD == "cmax") |>
  dplyr::left_join(reference, by = "treatment")
stopifnot(
  # Same parameters on both sides (Eq 6 plateau vs the ODE solve), so a tight
  # bound is correct; 211 is the paper's rounding of 210.8.
  all(abs(cmax_sim$PPORRES / cmax_sim$cmax - 1) < 0.002)
)
```

# Assumptions and deviations

- **Initial condition.** Eq 1 is printed without an initial condition. The
  maintainers set `central(0) <- TRIG * vc`, the measured baseline TG times
  the fixed plasma volume. Supplementary Figure S7 selects this choice: its
  early 90th percentile (~165 mg/dL at 0.1 h) is matched by the baseline
  start (~169 mg/dL) and not by a steady-state start `EndVLDL / Kel`
  (~217 mg/dL). After the infusion, TG relaxes to the model's own steady state
  `EndVLDL / (Kel x V)`, which differs from `TRIG` patient by patient.
  `TRIG` must therefore be the pre-infusion baseline, held constant.
- **Omega scale.** Table 4 prints "omega Kel = 43.3%". The maintainers read
  it as the SD of the exponential random effect (variance 0.433^2 = 0.1875),
  because it reproduces the paper's variability-propagation ratio
  (1.900 vs published 1.894 +/- 0.014). The CV reading gives 1.864.
- **Nomogram offset.** The published Figure 2 contours sit a uniform ~5.5%
  higher in baseline TG than Eq 7 evaluated with the Table 4 estimates
  (contour shape agrees to within a 2% log-SD over 26 digitised points). Eq 7
  with the Table 4 estimates reproduces the paper's own typical-patient rate
  (0.121 g/kg/h, Supplementary Table S6) and all four Table S6 sensitivity
  fits, so the packaged model follows Table 4 and Eq 7. The figure offset is
  recorded here as a deviation of the published drawing.
- **Units.** `amt` and `rate` are in mg of TG (1 g of emulsion fat = 1000 mg
  TG), the plasma volume is in dL and concentrations are in mg/dL, matching
  the paper's EndVLDL (mg/h) and V (dL).
- **Virtual cohort.** Body weight, baseline TG, infusion rate and infusion
  duration were drawn independently from piecewise-linear interpolations of
  the Table 1 quantiles. Correlations among them are not reported.
- **Covariate screen.** Apolipoprotein C-III, apolipoprotein C-II and serum
  albumin were significant on `Kel` in the univariate screen (Table 2), but
  none was retained once baseline TG was in the model (Table 3).
  Apolipoprotein C-III and albumin are listed in the model's
  `covariatesDataExcluded` metadata. Apolipoprotein C-II, which has no
  registered covariate column, is documented here only.
- **Scope.** The model assumes first-order TG elimination. The authors note
  that elimination appears saturated above about 1000 mg/dL (Supplementary
  Model Selection, Figure S2), so predictions in that range are
  underestimates by design.
- No erratum or correction notice for this article was found on Europe PMC
  as of 2026-10-02.
