---
title: "PI3K-inhibitor GI toxicity QSP model (Gadkar 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PI3K-inhibitor GI toxicity QSP model (Gadkar 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-load}
mod <- rxode2::rxode(readModelDb("Gadkar_2022_pi3kiColitis_qsp"))
```

* Citation: `r mod$reference`
* Description: `r mod$description`
* Article: <https://doi.org/10.1002/psp4.12749> (open access)

The paper describes the model in prose and its Table 1. The full equation
list is in Supplement S10 of the supporting information. Every parameter
value, dose object and per-drug parameter set comes from the authors'
deposited SimBiology project (`GI_risk_model.sbproj`, `Parameters.xlsx`
and `VPOP.mat`, supporting-information archive `PSP4-11-616-s001.zip`).
The maintainers read these files directly, and the reaction rates and
rules in the project are identical to the text of Supplement S10.

# Population

The model is a quantitative systems pharmacology (QSP) representation of
a cross-section of colonic mucosa (epithelium, lamina propria and
gut-associated lymphoid tissue) plus a blood compartment that supplies
immune cells. It was calibrated to clinical diarrhea and colitis
incidences for six PI3K inhibitors: taselisib, idelalisib, duvelisib,
umbralisib, alpelisib and pictilisib (paper Table 2). It was also
calibrated to individual diarrhea- and colitis-onset times from a
proprietary taselisib trial. Copanlisib was held out for validation.

Between-patient variability is represented by a prevalence-weighted
virtual population (VPop) of 704 virtual patients that vary 88
mechanistic parameters. This package carries only the authors' baseline
virtual patient (`VP0 (baseline VP)`). The full VPop is a set of 704
weighted parameter vectors, not a parametric distribution, so it cannot
be expressed as inter-individual variability. See the reproduction
section below for how the packaged equations perform against it.

```{r population}
str(mod$population)
```

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Drug PK (depot, central, peripheral1; F flux, lag, zero-order input) | taselisib defaults | Supplement S10 fluxes `PI3Ki_*`; Table S5; deposited variant `Taselisib (GDC-0032)` and dose `PI3Kinh_QD_4mg_Tablet` |
| `Cc = 1000 * central / vc` (ng/mL) | -- | S10 rule `Blood.PI3K_Inh_circ_total = PI3Kinh_circ*1e6/(PI3Kinh_Vc*1000)` |
| `Cu = fu * Cc`, `fu = 1 - 0.76` | 0.24 | S10 rule `Blood.PI3K_Inh_circ_conc`; Table S5 bound fraction |
| `Cmucosa = kp_gut * Cu`, `kp_gut` | 0.5 | S10 rule `Mucosa.PI3K_Inhibitor`; deposited `PI3Kinh_partition` |
| Isoform activity `1 - imax * C / (ic50 + C)` | -- | S10 rules `PI3Ka_EC`, `PI3Kd_APC`, ... (identical per isoform across cell types) |
| `lic50_pi3ka/b/g/d` (unbound, ng/mL) | 0.13 / 4 / 0.45 / 0.055 | Deposited taselisib variant; Table S3 (nM) |
| Mucosal fluxes (recruitment, activation, differentiation, proliferation, cytokine production and clearance) | -- | Supplement S10 'Fluxes' |
| `Epithelial_Quality = min(1, EChealthy + 0.2 * ECact)` | -- | S10 'Repeated Assignments'; Table 1 |
| `GI_Risk_Score` | weights 0.2 / 0.8 | S10 'Repeated Assignments'; Table 1 |
| `d/dt(RISK) = k_healing * (GI - RISK * (GI < threshold))` | -- | S10 fluxes `insult`, `healing` |
| 21 ODEs | -- | Supplement S10 'ODEs' |
| 155 mechanistic parameters | see model file | `Parameters.xlsx` tab 'All parameters' (= variant `VP0 (baseline VP)`); rounded values in Table S1 |
| Diarrhea criterion `max(GI_Risk_Score) > 0.6` | 0.6 | Table 1; `getAE.m` `threshold0` |
| Colitis criterion `max(RISK) > 200` | 200 | `getAE.m` `threshold2` (Table 1 prints 120; see Assumptions) |
| Initial conditions | drug-free steady state | Converged run-in of the deposited procedure (`simulate_Vpop.m`) |

# Per-drug parameters

The seven inhibitors share every mechanistic parameter. They differ only
in their PK, protein binding and isoform IC50s, which the deposited
project stores as one SimBiology variant per drug. The IC50s are unbound
concentrations in ng/mL, which is Table S3 in nM times the molecular
weight divided by 1000.

```{r drugs}
drugs <- tibble::tribble(
  ~drug,        ~cl,   ~vc,    ~q,     ~vp,   ~ka,   ~tlag, ~fb,    ~ic50a,   ~ic50b, ~ic50g, ~ic50d, ~amt, ~ii, ~route,
  "Taselisib",   5,     200,    10,     50,    0.7,   0.2,   0.76,   0.13,     4,      0.45,   0.055,  4,    24,  "oral",
  "Idelalisib",  14.88, 22.65,  11.82,  72.97, 0.482, 0.247, 0.84,   340.64,   234.71, 36.97,  1.04,   150,  12,  "oral",
  "Duvelisib",   5,     200,    10,     50,    0.7,   0.2,   0.7625, 667.81,   35.43,  11.26,  1.04,   25,   12,  "oral",
  "Umbralisib",  8.7,   550,    10,     160,   0.8,   0.2,   0.7625, 101200,   506,    485,    10.12,  800,  24,  "oral",
  "Alpelisib",   11.5,  118,    0,      1,     0.784, 0.489, 0.9,    2.21,     510,    128,    110,    300,  24,  "oral",
  "Pictilisib",  34.12, 441.4,  8.145,  228.1, 2.065, 0.2,   0.95,   1.38,     37.17,  22.02,  0.83,   340,  24,  "oral",
  "Copanlisib",  38.4,  317,    113,    815,   0.7,   0.2,   0.84,   0.276,    2.05,   3.54,   0.387,  60,   NA,  "iv"
)
drugs |>
  dplyr::rename(
    "Drug" = drug, "CL (L/h)" = cl, "Vc (L)" = vc, "Q (L/h)" = q, "Vp (L)" = vp,
    "ka (1/h)" = ka, "Lag (h)" = tlag, "Bound fraction" = fb,
    "IC50 alpha" = ic50a, "IC50 beta" = ic50b, "IC50 gamma" = ic50g, "IC50 delta" = ic50d,
    "Dose (mg)" = amt, "Interval (h)" = ii, "Route" = route
  ) |>
  knitr::kable(caption = "Deposited per-drug parameter sets (IC50s are unbound, ng/mL).")

# Named parameter vector for one drug, passed to rxSolve(params = ...).
# Alpelisib is one-compartment in the deposit (Q = 0); a log() of zero
# is not allowed, so Q is set to 1e-10 L/h, which is numerically identical.
drug_params <- function(d) {
  c(
    lcl = log(d$cl), lvc = log(d$vc), lq = log(max(d$q, 1e-10)), lvp = log(d$vp),
    lka = log(d$ka), ltlag = log(d$tlag), fu = 1 - d$fb,
    lic50_pi3ka = log(d$ic50a), lic50_pi3kb = log(d$ic50b),
    lic50_pi3kg = log(d$ic50g), lic50_pi3kd = log(d$ic50d)
  )
}

# Copanlisib: 60 mg IV on days 1, 8 and 15 of each 28-day cycle (deposited
# dose 'Copanlisib_60mg').
cop_times <- as.vector(outer(c(0, 168, 336), seq(0, 9) * 672, "+"))

# Dosing events for one drug from time 0 over `days` days. Oral doses carry
# rate = -2 so rxode2 uses the modelled 0.1-h input duration (dur(depot)).
drug_events <- function(d, days, obs) {
  ev <- if (d$route == "oral") {
    rxode2::et(amt = d$amt, ii = d$ii, addl = days * 24 / d$ii - 1, cmt = "depot", rate = -2)
  } else {
    rxode2::et(amt = d$amt, time = cop_times[cop_times < days * 24], cmt = "central")
  }
  rxode2::et(ev, obs)
}
```

# Drug-free steady state

The initial conditions are the drug-free steady state of the baseline
virtual patient. Without drug, every state should stay where it starts.
This is the paper's first qualification criterion: an untreated baseline
with a stable, healthy epithelium.

```{r steady-state}
ss <- rxode2::rxSolve(mod, rxode2::et(seq(0, 24 * 240, by = 24 * 10)), returnType = "data.frame")
states <- c(
  "T_Eff_act", "APC_act", "Treg_act", "APC_inact", "T_Eff_inact", "Treg_inact",
  "T_quies", "Neutrophils", "EChealthy", "ECact", "nTreg_circ", "RISK"
)
drift <- sapply(states, function(s) max(abs(ss[[s]] / ss[[s]][1] - 1)))
signif(drift, 2)
# Numerical drift only (LSODA rtol 1e-6); a mis-transcribed rate constant
# moves at least one state by percent-level amounts within 240 days.
stopifnot(all(drift < 1e-4))
c(
  Epithelial_Quality = ss$Epithelial_Quality[1],
  GI_Risk_Score = ss$GI_Risk_Score[1]
)
```

The steady-state epithelial quality is 0.982. Table S1 lists 0.98
("assumed to be close to 100% in healthy conditions"). Table S1 lists
the baseline GI risk score as 0.23. The model's steady state is 0.021,
which matches 0.023 to two figures, so 0.23 is most likely a misplaced
decimal. A value of 0.23 would also put the healthy colon at the healing
threshold of 0.25.

# Replicate Figure 2: isoform activity profiles

Figure 2 plots the normalized activity of the four isoforms in the
mucosa over the first 20 days of each drug's regimen. These curves
depend only on PK and IC50s, not on the virtual patient, so the
packaged model must reproduce them exactly.

```{r fig2, fig.height = 6}
obs20 <- seq(0, 20 * 24, by = 1)
act <- dplyr::bind_rows(lapply(seq_len(nrow(drugs)), function(i) {
  d <- drugs[i, ]
  s <- rxode2::rxSolve(mod, drug_events(d, 21, obs20), params = drug_params(d), returnType = "data.frame")
  data.frame(
    drug = d$drug, time = s$time,
    alpha = s$PI3Ka_act, beta = s$PI3Kb_act, gamma = s$PI3Kg_act, delta = s$PI3Kd_act
  )
}))
act |>
  tidyr::pivot_longer(alpha:delta, names_to = "isoform", values_to = "activity") |>
  dplyr::mutate(drug = factor(drug, levels = drugs$drug)) |>
  ggplot(aes(time / 24, activity, colour = isoform)) +
  geom_line() +
  facet_wrap(~drug, ncol = 4) +
  scale_colour_manual(values = c(alpha = "red", beta = "cyan3", gamma = "magenta", delta = "blue")) +
  labs(x = "Time (days)", y = "Activity (fraction)", colour = "PI3K isoform") +
  theme_bw()
```

Replicates Figure 2 of Gadkar et al. (2022): normalized mucosal PI3K
pathway activity (alpha red, beta cyan, gamma magenta, delta blue).

Table 2 of the paper puts each drug's average alpha and delta inhibition
into a category. The table below computes the mean inhibition over the
third week (days 14 to 21) and checks it against those categories.

```{r table2-categories}
inh <- act |>
  dplyr::filter(time >= 14 * 24) |>
  dplyr::group_by(drug) |>
  dplyr::summarise(alpha_inh = 100 * (1 - mean(alpha)), delta_inh = 100 * (1 - mean(delta))) |>
  dplyr::mutate(drug = factor(drug, levels = drugs$drug)) |>
  dplyr::arrange(drug)
paper <- data.frame(
  drug = drugs$drug,
  alpha_paper = c(
    "Strong (>90%)", "Weak (10-25%)", "Weak (~5%)", "Negligible (~0%)",
    "Strong (>90%)", "Strong (75-90%)", "Strong->weak (90->15%)"
  ),
  delta_paper = c(
    "Strong (>90%)", "Strong (>90%)", "Strong (>90%)", "Strong (>90%)",
    "Mod/weak (15-55%)", "Strong (80-95%)", "Strong->weak (90->15%)"
  )
)
inh |>
  dplyr::mutate(drug = as.character(drug)) |>
  dplyr::left_join(paper, by = "drug") |>
  dplyr::mutate(alpha_inh = round(alpha_inh, 1), delta_inh = round(delta_inh, 1)) |>
  dplyr::rename(
    "Drug" = drug, "Mean alpha inhibition (%)" = alpha_inh, "Mean delta inhibition (%)" = delta_inh,
    "Paper alpha category" = alpha_paper, "Paper delta category" = delta_paper
  ) |>
  knitr::kable(caption = "Mean mucosal inhibition over days 14-21 versus Table 2 of the paper.")
# Deterministic (no IIV): the inhibition follows from PK and IC50 alone.
# The ranges below are the paper's Table 2 categories; for the "~"
# categories they are widened to +/- 5 points around the stated value.
# Copanlisib's category (strong -> weak, 90 -> 15%) describes the swing
# between a weekly dose and the end of that week, checked separately below.
g <- function(d, col) inh[[col]][inh$drug == d]
stopifnot(
  g("Taselisib", "alpha_inh") > 90, g("Taselisib", "delta_inh") > 90,
  g("Idelalisib", "alpha_inh") > 10, g("Idelalisib", "alpha_inh") < 25, g("Idelalisib", "delta_inh") > 90,
  g("Duvelisib", "alpha_inh") < 10, g("Duvelisib", "delta_inh") > 90,
  g("Umbralisib", "alpha_inh") < 5, g("Umbralisib", "delta_inh") > 90,
  g("Alpelisib", "alpha_inh") > 90, g("Alpelisib", "delta_inh") > 15, g("Alpelisib", "delta_inh") < 55,
  g("Pictilisib", "alpha_inh") > 75, g("Pictilisib", "alpha_inh") < 90,
  g("Pictilisib", "delta_inh") > 80, g("Pictilisib", "delta_inh") < 95
)

# Copanlisib: the week after the day-14 dose (days 14-20).
cop <- act |> dplyr::filter(drug == "Copanlisib", time >= 14 * 24, time <= 20 * 24)
cop_swing <- c(
  alpha_peak_inh = 100 * (1 - min(cop$alpha)), alpha_trough_inh = 100 * (1 - max(cop$alpha)),
  delta_peak_inh = 100 * (1 - min(cop$delta)), delta_trough_inh = 100 * (1 - max(cop$delta))
)
round(cop_swing, 1)
stopifnot(
  cop_swing[["alpha_peak_inh"]] > 85, cop_swing[["alpha_trough_inh"]] < 20,
  cop_swing[["delta_peak_inh"]] > 85, cop_swing[["delta_trough_inh"]] < 20
)
```

# Replicate Figure 4a and 4d: IC50 sensitivity for taselisib

The paper raises the taselisib alpha IC50 (nominal 0.29 nM) and delta
IC50 (nominal 0.12 nM) 10-, 30- and 100-fold. It reports that alpha
inhibition falls "from greater than 95% to ~20% on average" and delta
inhibition "from greater than 95% to ~40%" at 100-fold.

```{r fig4}
tas <- drugs[drugs$drug == "Taselisib", ]
sens <- dplyr::bind_rows(lapply(c(1, 10, 30, 100), function(f) {
  dplyr::bind_rows(lapply(c("alpha", "delta"), function(iso) {
    p <- drug_params(tas)
    nm <- if (iso == "alpha") "lic50_pi3ka" else "lic50_pi3kd"
    p[[nm]] <- p[[nm]] + log(f)
    s <- rxode2::rxSolve(mod, drug_events(tas, 21, obs20), params = p, returnType = "data.frame")
    data.frame(
      fold = f, isoform = iso, time = s$time,
      activity = if (iso == "alpha") s$PI3Ka_act else s$PI3Kd_act
    )
  }))
}))
ggplot(sens, aes(time / 24, activity, colour = factor(fold))) +
  geom_line() +
  facet_wrap(~isoform, labeller = labeller(isoform = c(alpha = "PI3K-alpha (Fig 4a)", delta = "PI3K-delta (Fig 4d)"))) +
  labs(x = "Time (days)", y = "Activity (fraction)", colour = "IC50 fold") +
  theme_bw()
sens_inh <- sens |>
  dplyr::filter(time >= 14 * 24) |>
  dplyr::group_by(isoform, fold) |>
  dplyr::summarise(inhibition = 100 * (1 - mean(activity)), .groups = "drop")
knitr::kable(sens_inh, digits = 1, caption = "Mean inhibition (%) over days 14-21 by IC50 fold.")
inh_at <- function(iso, f) sens_inh$inhibition[sens_inh$isoform == iso & sens_inh$fold == f]
stopifnot(
  inh_at("alpha", 1) > 95, inh_at("delta", 1) > 95,
  abs(inh_at("alpha", 100) - 20) < 5,
  abs(inh_at("delta", 100) - 40) < 5
)
```

Replicates Figure 4a and 4d of Gadkar et al. (2022).

# Replicate Figure 6a: taselisib dosing holidays

```{r fig6a, fig.height = 6}
d42 <- seq(0, 41) * 24
regimens <- list(
  "Daily" = d42,
  "5 days on, 2 off" = d42[(d42 / 24) %% 7 < 5],
  "14 days on, 7 off" = d42[(d42 / 24) %% 21 < 14]
)
hol <- dplyr::bind_rows(lapply(names(regimens), function(r) {
  ev <- rxode2::et(amt = tas$amt, time = regimens[[r]], cmt = "depot", rate = -2) |>
    rxode2::et(seq(0, 42 * 24, by = 1))
  s <- rxode2::rxSolve(mod, ev, params = drug_params(tas), returnType = "data.frame")
  data.frame(
    regimen = r, time = s$time,
    alpha = s$PI3Ka_act, beta = s$PI3Kb_act, gamma = s$PI3Kg_act, delta = s$PI3Kd_act
  )
}))
hol |>
  tidyr::pivot_longer(alpha:delta, names_to = "isoform", values_to = "activity") |>
  dplyr::mutate(regimen = factor(regimen, levels = names(regimens))) |>
  ggplot(aes(time / 24, activity, colour = regimen)) +
  geom_line() +
  facet_wrap(~isoform) +
  labs(x = "Time (days)", y = "Activity (fraction)", colour = "Regimen") +
  theme_bw()
```

Replicates Figure 6a of Gadkar et al. (2022). As the paper notes, the
alpha, gamma and delta activities do not return to 1 even during the
7-day holiday.

```{r fig6a-check}
holiday <- hol |>
  dplyr::filter(regimen == "14 days on, 7 off", time > 14 * 24, time <= 21 * 24)
c(max_alpha = max(holiday$alpha), max_gamma = max(holiday$gamma), max_delta = max(holiday$delta))
stopifnot(max(holiday$alpha) < 1, max(holiday$gamma) < 1, max(holiday$delta) < 1)
```

# Replicate Figure S1: response to an acute epithelial injury

Figure S1 shows the response to an acute epithelial injury given on
day 10, using the deposited dose `Epi_activation_single` (500 units into
`Activating_Event`). Neutrophils, APCs and the GI risk score rise and
then resolve, "within 20-30 days consistent with chemically induced
colitis in preclinical models".

```{r figS1, fig.height = 5.5}
ev_inj <- rxode2::et(amt = 500, time = 240, cmt = "Activating_Event", rate = -2) |>
  rxode2::et(seq(0, 70 * 24, by = 4))
inj <- rxode2::rxSolve(mod, ev_inj, returnType = "data.frame")
inj |>
  dplyr::select(time, Epithelial_Quality, GI_Risk_Score, APC_act, Neutrophils) |>
  tidyr::pivot_longer(-time) |>
  ggplot(aes(time / 24, value)) +
  geom_line() +
  facet_wrap(~name, scales = "free_y") +
  labs(x = "Time (days)", y = NULL) +
  theme_bw()
# Remaining excursion above the pre-injury level, as a fraction of the peak
# excursion, 30 days after the injury (day 40).
excursion <- function(x) (x[inj$time == 40 * 24] - x[1]) / (max(x) - x[1])
res_s1 <- c(
  peak_GI = max(inj$GI_Risk_Score),
  peak_neutrophils = max(inj$Neutrophils),
  GI_left_day40 = excursion(inj$GI_Risk_Score),
  APC_left_day40 = excursion(inj$APC_act),
  neutrophils_left_day40 = excursion(inj$Neutrophils)
)
signif(res_s1, 3)
# Deterministic. The injury must produce a real transient (the GI risk
# score rises more than 10-fold), and less than 5% of each excursion may
# remain 30 days later (the paper: resolved within 20-30 days).
stopifnot(
  res_s1[["peak_GI"]] > 10 * inj$GI_Risk_Score[1],
  res_s1[["GI_left_day40"]] < 0.05,
  res_s1[["APC_left_day40"]] < 0.05,
  res_s1[["neutrophils_left_day40"]] < 0.05
)
```

In the baseline virtual patient, almost all of the inflammation
resolves 26-30 days after the injury. A slow tail in the APC and T-cell
densities takes several more weeks to return fully to baseline.

Replicates Figure S1 of Gadkar et al. (2022) for the baseline virtual
patient. The paper plots the VPop mean and 5-95% band.

# Baseline virtual patient on taselisib

The time course of one virtual patient on taselisib 4 mg once daily
(compare Figure S8 for the VPop). The baseline virtual patient is a
patient who does not develop diarrhea or colitis.

```{r vp0-taselisib, fig.height = 5.5}
obs240 <- seq(0, 240 * 24, by = 6)
vp0 <- rxode2::rxSolve(mod, drug_events(tas, 240, obs240), params = drug_params(tas), returnType = "data.frame")
vp0 |>
  dplyr::select(time, Epithelial_Quality, GI_Risk_Score, APC_act, Neutrophils, T_Eff_act, RISK) |>
  tidyr::pivot_longer(-time) |>
  ggplot(aes(time / 24, value)) +
  geom_line() +
  facet_wrap(~name, scales = "free_y") +
  labs(x = "Time (days)", y = NULL) +
  theme_bw()
c(max_GI_Risk_Score = max(vp0$GI_Risk_Score), max_RISK = max(vp0$RISK))
# Deterministic: the baseline virtual patient stays below both onset
# criteria (diarrhea GI > 0.6, colitis RISK > 200) under taselisib.
stopifnot(max(vp0$GI_Risk_Score) < 0.6, max(vp0$RISK) < 200)
```

# Reproduction of the virtual-population incidences

The paper's headline numbers are prevalence-weighted incidences across
the 704-patient VPop. These are computed as follows:

* Run two 240-day drug-free run-ins.
* Dose for 240 days.
* Count diarrhea when `max(GI_Risk_Score) > 0.6` and colitis when
  `max(RISK) > 200`.

On 2026-10-01 the maintainers re-simulated the authors' deposited
`VPOP.mat` (88 parameters x 704 weighted patients) with the equations
and parameters in this model file. The results are below. They are not
recomputed at render time because the VPop is not distributed with the
package.

| Scenario | Paper diarrhea | Re-simulated diarrhea | Paper colitis | Re-simulated colitis |
|----------|---------------:|----------------------:|--------------:|---------------------:|
| Taselisib 4 mg QD (Tables 2, 3) | 52% | 51.9% | 10.2% | 10.2% |
| Idelalisib 150 mg BID (Table 2) | -- | 40.1% | 9.5% | 9.5% |
| Duvelisib 25 mg BID (Table 2) | -- | 29.7% | 11.4% | 11.4% |
| Umbralisib 800 mg QD (Table 2) | -- | 40.2% | 1.1% | 1.1% |
| Alpelisib 300 mg QD (Table 2) | -- | 31.7% | 2.4% | 2.4% |
| Pictilisib 340 mg QD (Table 2) | -- | 71.1% | 16.3% | 16.3% |
| Copanlisib 60 mg IV d1/8/15 q28d (Table 2) | -- | 22.2% | 1.5% | 1.5% |
| Taselisib 5 days on, 2 off (Table 3) | 53% | 52.4% | 15% | 14.9% |
| Taselisib 14 days on, 7 off (Table 3) | 58% | 56.5% | 15% | 14.3% |
| Taselisib, alpha IC50 x100 (Figure 4b, c) | 27% | 26.7% | 7% | 6.4% |
| Taselisib, delta IC50 x100 (Figure 4e, f) | 18% | 17.5% | 3% | 2.5% |

All seven Table 2 colitis incidences reproduce to the printed decimal.
Two of the 704 virtual patients could not be integrated by LSODA on the
copanlisib schedule; they are counted as event-free. The small
differences on the dosing-holiday regimens are consistent with the exact
on/off day boundaries, which the deposit does not include.

# PKNCA validation of the PK layer

The paper reports no NCA. The PK layer is therefore checked against its
own closed form. At steady state, the AUC over one dosing interval
equals `F * Dose / CL`, so `1000 * Dose / CL` in ng*h/mL with F = 1.
The check uses each oral regimen after 28 days of dosing, which is at
least 11 half-lives for the slowest drug (umbralisib).

```{r pknca}
oral <- drugs[drugs$route == "oral", ]
conc <- dplyr::bind_rows(lapply(seq_len(nrow(oral)), function(i) {
  d <- oral[i, ]
  t0 <- 28 * 24
  obs <- t0 + seq(0, d$ii, by = 0.05)
  ev <- rxode2::et(amt = d$amt, ii = d$ii, addl = (t0 + d$ii) / d$ii, cmt = "depot", rate = -2) |>
    rxode2::et(obs)
  s <- rxode2::rxSolve(mod, ev, params = drug_params(d), returnType = "data.frame", rtol = 1e-10, atol = 1e-12)
  data.frame(id = 1L, treatment = d$drug, time = s$time - t0, Cc = s$Cc)
}))
conc <- conc |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(Cc = pmax(Cc, 0))
dose_df <- data.frame(id = 1L, treatment = oral$drug, time = 0, amt = oral$amt)
intervals <- data.frame(
  start = 0, end = oral$ii, treatment = oral$drug,
  auclast = TRUE, cmax = TRUE, tmax = TRUE, cmin = TRUE
)
nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc, Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id),
  intervals = intervals
))
res <- as.data.frame(nca$result) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::left_join(dplyr::transmute(oral, treatment = drug, auc_closed = 1000 * amt / cl), by = "treatment") |>
  dplyr::mutate(pct_diff = 100 * (auclast / auc_closed - 1))
res |>
  dplyr::rename(
    "Drug" = treatment, "Cmax,ss (ng/mL)" = cmax, "Tmax (h)" = tmax, "Cmin,ss (ng/mL)" = cmin,
    "AUCtau,ss PKNCA (ng*h/mL)" = auclast, "Dose/CL (ng*h/mL)" = auc_closed, "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = 3, caption = "Steady-state NCA of the PK layer versus the closed-form AUC.")
# Linear-up / log-down trapezoid on a 0.05-h grid. Measured residuals are
# below 0.03%; the largest, umbralisib's, is its remaining approach to
# steady state after 28 days. A wrong CL, dose or unit factor moves this
# by far more than 1%.
stopifnot(all(abs(res$pct_diff) < 1))
```

# Assumptions and deviations

* **Default drug and virtual patient.** The packaged parameters are the
  deposited baseline virtual patient `VP0 (baseline VP)` with taselisib
  PK and potency, which is the variant the deposited project applies by
  default. Other drugs are simulated by overriding the 11 drug
  parameters (see `drug_params()` above and `population$notes`).
* **Virtual population not carried.** The 704 prevalence-weighted
  virtual patients are a non-parametric set, not an IIV distribution.
  The model is deterministic and has no residual error, because the
  paper fits none.
* **Initial conditions.** The deposit starts from the Table S1 cell
  densities with no epithelium and runs two 240-day drug-free run-ins
  before each treatment. The packaged initial conditions are the
  converged end of that run-in for the baseline virtual patient. If
  mechanistic parameters are changed, simulate a drug-free run-in (as
  the reproduction above did) before dosing.
* **Colitis threshold.** Table 1 defines colitis onset as the AUC of
  the GI risk score crossing 120 normalized-units-day. The deposited
  code instead tests the state `RISK`, which accumulates
  `k_healing * GI` per hour and heals while GI is below 0.25, against
  200. Only the deposited criterion reproduces Tables 2 and 3, so it is
  used here.
* **Baseline GI risk score.** Table S1 lists 0.23, but the model's
  steady state is 0.021 (see the steady-state section).
* **IC50 units and Table S3.** The deposited IC50s are unbound ng/mL.
  For most drugs they equal the Table S3 nM values times the molecular
  weight. The taselisib beta IC50 of 4 ng/mL corresponds to the
  footnoted "30x relative selectivity" (0.29 x 30 = 8.7 nM), not to the
  0.9 nM printed in the table. Pictilisib's deposited IC50s do not
  follow Table S3 by a single factor. The deposited values, which
  generate the published results, are used throughout.
* **Collapsed activity rules.** The deposit computes the same isoform
  activity separately for each cell type, e.g. `PI3Ka_EC`, `PI3Ka_APC`
  and `PI3Ka_T_Eff`. These are identical expressions of the mucosal
  concentration, so the model carries one activity per isoform in the
  mucosa (`PI3Ka_act` and so on) and two in blood (`PI3Kg_act_circ` and
  `PI3Kd_act_circ`). The deposited rule `PI3Kg_Neutro` and the
  `PI3K_Inh_Tissue_SS` average feed no rate and are omitted.
* **Unused deposited parameters.** Twelve parameters in `Parameters.xlsx`
  appear in no rate law or rule of the deposit. They describe an
  immature-epithelial-cell pool and alternative epithelial-quality
  weights that are not part of the final model, and they are omitted:
  `Epi_imm_qual_weight`, `Epi_Cell_act_weight`,
  `Epi_quies_qual_weight`, `Inflam_Colitis_w`, `Epi_imm_diff_rate_k`,
  `PI3Ka_ECimm_diff_EC50`, `PI3Ka_ECimm_diff_Emax`,
  `Epi_proliferation_rate_k`, `PI3Ka_ECimm_apop_IC50`,
  `PI3Ka_ECimm_apop_Imax`, `Epi_imm_apoptosis_rate_k` and
  `Epi_act_rate_k`.
* **Oral input.** Deposited oral doses enter the gut compartment as a
  0.1-h zero-order input after a drug-specific lag. Use `rate = -2` on
  oral dose records so the modelled duration is applied. Copanlisib is
  an IV bolus into `central`.
* **Acute-injury timing.** The deposited `Epi_activation_single` dose
  starts at time 10. Figure S1 places the injury on day 10, so the
  replication doses at 240 h.
