---
title: "Niraparib (Gaffney 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Niraparib (Gaffney 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Gaffney_2026_niraparib"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1002/jcph.70210>

Niraparib is an oral poly(ADP-ribose) polymerase (PARP) inhibitor approved for
maintenance treatment of advanced ovarian cancer. Gaffney and colleagues updated
an earlier niraparib population PK model by adding intensively sampled phase 1
data from two studies (HEPATIC and TABLET), and used those data to re-derive the
absorption model. The result -- which the authors call the *niraparib reference
population PK model* -- keeps the earlier three-compartment linear-elimination
disposition but replaces the earlier zero-order-release-plus-lag-time absorption
with a three-transit-compartment chain.

# Population

The analysis pooled six studies and 14,106 quantifiable plasma niraparib
concentrations from 1686 patients (Table 1):

```{r population-table, echo = FALSE}
tibble::tribble(
  ~Study,     ~Phase, ~`Cancer type`,                       ~Patients, ~Observations,
  "PN001",    "1",    "Advanced solid tumours",                   104L,  2096L,
  "NOVA",     "3",    "Recurrent ovarian cancer",                 405L,  2054L,
  "QUADRA",   "2",    "Advanced relapsed ovarian cancer",         455L,  1410L,
  "PRIMA",    "3",    "First-line advanced ovarian cancer",       480L,  1856L,
  "HEPATIC",  "1",    "Advanced solid tumours",                    17L,   203L,
  "TABLET",   "1",    "Advanced solid tumours",                   225L,  6487L
) |>
  knitr::kable(caption = "Studies pooled into the niraparib reference model (Gaffney 2026 Table 1).")
```

Most patients were female (1544, 91.6%) and White (1457, 86.4%); 1410 (83.6%)
had ovarian cancer and 1459 (86.5%) had normal hepatic function. HEPATIC and
TABLET are the two studies added by this analysis: HEPATIC enrolled 17 patients
(8 with moderate hepatic impairment, 9 with normal hepatic function) given a
single 300 mg dose, and TABLET compared tablet with capsule formulations and
assessed a high-fat meal after a single 300 mg tablet dose. TABLET alone
contributed 46% of the analysis observations.

The same information is available programmatically from the model's `population`
metadata:

```{r population-meta}
str(ui$population, max.level = 1, give.attr = FALSE)
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Gaffney_2026_niraparib.R` carries an in-file comment
naming its source location. They are collected here for review.

| Equation / parameter | Value | Source location |
|---|---|---|
| Structure: 3 transit compartments -> central, 3-compartment disposition | n/a | Figure 3; Base Population PK Model |
| Continuous covariate form `P = theta_k * (X / M(X))^theta_j` | n/a | Covariate Analysis, page 4 |
| Categorical covariate form `P = theta_k * (1 + theta_j)^X` | n/a | Covariate Analysis, page 4 |
| `lcl` (CL/F) | 15.9 L/h | Table 3 |
| `lvc` (Vc/F) | 450 L | Table 3 |
| `lq` (Q1/F) | 43.8 L/h | Table 3 |
| `lvp` (Vp1/F) | 395 L | Table 3 |
| `lq2` (Q2/F) | 2.19 L/h | Table 3 |
| `lvp2` (Vp2/F) | 361 L | Table 3 |
| `lmtt` (MTT) | 1.78 h | Table 3 |
| `lfdepot` (F1) | 1, fixed | Table 3; Base Population PK Model |
| `e_alb_cl` | 0.742 | Table 3 (CL/F ~ ALBBL exponent) |
| `e_alp_cl` | -0.074 | Table 3 (CL/F ~ ALPBL exponent) |
| `e_crcl_cl` | 0.287 | Table 3 (CL/F ~ CrCLBL exponent) |
| `e_alb_vc` | 0.363 | Table 3 (Vc/F ~ ALBBL exponent) |
| `e_wt_vc` | 0.577 | Table 3 (Vc/F ~ WTBL exponent) |
| `e_alb_vp` | 1.02 | Table 3 (Vp1/F ~ ALBBL exponent) |
| `e_fed_fdepot` | 0.236 | Table 3 (F1 ~ fed fractional change) |
| `e_fed_mtt` | 1.15 | Table 3 (MTT ~ fed fractional change) |
| Covariate reference values ALB 4 g/dL, ALP 83 IU/L, CrCL 82.37 mL/min, WT 70 kg | n/a | Figure 5 legend; Results |
| `etalcl`, `etalvc`, `etalmtt`, `etalvp2`, `etalfdepot` | 24.0, 23.9, 56.1, 53.8, 33.8 %CV | Table 3 Random Effects |
| `expSd` and the five study-specific magnitudes | 0.227, 0.350, 0.381, 0.451, 0.324, 0.146 | Table 3 Residual Error |

```{r ini-table, echo = FALSE}
ui$iniDf |>
  dplyr::filter(!is.na(ntheta)) |>
  dplyr::transmute(Parameter = name, Estimate = signif(est, 6),
                   Fixed = fix, Label = label) |>
  knitr::kable(caption = "Packaged `ini()` values, read back from the model object.")
```

## Reading the absorption chain

Table 3 reports a mean transit time (MTT) but no absorption rate constant, so the
single rate constant the paper labels `Ka` has to be derived from MTT. Figure 3
draws the final absorption model as

```
Oral dose -> (Transit) -Ka-> (Transit) -Ka-> (Transit) -Ka-> Central
```

with no lag time, no zero-order release and no separate depot: exactly **three**
first-order steps carrying the dose from the dosing record to the central
compartment, all at the same rate. The arrival time in `central` is therefore
Erlang-distributed with shape 3 and rate `ktr`, whose mean -- the mean transit
time -- is `3 / ktr`. The model encodes `ktr <- 3 / mtt`, i.e. 1.685 1/h at the
typical MTT of 1.78 h. See *Assumptions and deviations* for the alternative
reading that was considered and rejected.

In the packaged file the three boxes are written as `depot`, `transit1` and
`transit2`, so the dose lands in the canonical extravascular dosing compartment;
the three `ktr` arrows are `depot -> transit1`, `transit1 -> transit2` and
`transit2 -> central`, matching Figure 3 arrow for arrow. This is the same
depot-plus-(n-1)-transit idiom used by `Comisar_2025_rimegepant`.

# Typical-value scenarios

The strongest validation available for this paper is a set of published
steady-state exposure ratios. In the Results the authors report, for each
covariate, the AUC ratio relative to a reference patient (ALB 4 g/dL, ALP
83 IU/L, CrCL 82.37 mL/min, WT 70 kg, fasted/unknown) at the observed minimum,
5th percentile, 95th percentile and maximum covariate values. Because AUC after
an oral dose is `F1 * Dose / (CL/F)` and none of these covariates acts on both
F1 and CL/F, each published ratio is a direct, deterministic test of one
covariate coefficient *and* its reference value.

```{r scenarios}
mod <- readModelDb("Gaffney_2026_niraparib")

ref_cov <- list(
  ALB = 40, ALP = 83, CRCL = 82.37, WT = 70, FED = 0,
  STUDY_NOVA = 0, STUDY_QUADRA = 0, STUDY_PRIMA = 0,
  STUDY_TABLET = 0, STUDY_HEPATIC = 0
)

scen <- tibble::as_tibble(ref_cov) |> dplyr::mutate(scenario = "Reference")
vary <- function(label, column, value) {
  out <- scen[1, ]
  out$scenario <- label
  out[[column]] <- value
  out
}

scen <- dplyr::bind_rows(
  scen,
  # ALB is held in the canonical SI unit g/L; the paper's g/dL values are x10.
  vary("ALB 1.7 g/dL (min)",  "ALB",  17),
  vary("ALB 3.1 g/dL (P5)",   "ALB",  31),
  vary("ALB 4.7 g/dL (P95)",  "ALB",  47),
  vary("ALB 7.9 g/dL (max)",  "ALB",  79),
  vary("CrCL 27.6 (min)",     "CRCL", 27.6),
  vary("CrCL 44.3 (P5)",      "CRCL", 44.3),
  vary("CrCL 147.4 (P95)",    "CRCL", 147.4),
  vary("CrCL 298.6 (max)",    "CRCL", 298.6),
  vary("ALP 0.9 (min)",       "ALP",  0.9),
  vary("ALP 49 (P5)",         "ALP",  49),
  vary("ALP 211.8 (P95)",     "ALP",  211.8),
  vary("ALP 1814 (max)",      "ALP",  1814),
  vary("Fed",                 "FED",  1)
) |>
  dplyr::mutate(id = dplyr::row_number())

# The terminal disposition half-life is ~136 h, so the observation window runs
# to 1000 h (about 7 terminal half-lives) to keep the AUC extrapolation small.
scen_times <- sort(unique(c(
  seq(0, 24, by = 0.05), seq(24, 96, by = 0.5), seq(96, 1000, by = 4)
)))

expand_events <- function(subjects, times, dose = 300) {
  subjects |>
    dplyr::group_split(id) |>
    lapply(function(s) {
      n <- length(times)
      dplyr::bind_cols(
        s[rep(1L, n + 1L), setdiff(names(s), c("time", "amt", "evid", "cmt"))],
        tibble::tibble(
          time = c(0, times),
          amt  = c(dose, rep(NA_real_, n)),
          evid = c(1L, rep(0L, n)),
          # Observations are placed on the ODE state `central`; rxode2 returns
          # the algebraic observable Cc as a column at those rows.
          cmt  = c("depot", rep("central", n))
        )
      )
    }) |>
    dplyr::bind_rows() |>
    as.data.frame()
}

scen_events <- expand_events(scen, scen_times)
stopifnot(!anyDuplicated(unique(scen_events[, c("id", "time", "evid")])))
```

```{r scenario-sim}
# zeroRe() gives the typical-value prediction the published ratios describe.
scen_sim <- rxode2::rxSolve(
  rxode2::zeroRe(mod), events = scen_events,
  keep = "scenario", omega = NA
) |>
  as.data.frame()

# A transit chain that silently solves to zero is a known rxode2 failure mode;
# this assertion is the cheap standing guard against it.
stopifnot(max(scen_sim$Cc, na.rm = TRUE) > 1, all(scen_sim$Cc >= 0, na.rm = TRUE))
```

Two structural identities can be checked before any NCA is run. The paper's
Discussion states that the total apparent volume of distribution
(Vc/F + Vp1/F + Vp2/F) is 1206 L, and the derived transit rate constant must
follow from the reported MTT:

```{r structural-identities}
tv <- scen_sim |> dplyr::filter(scenario == "Reference") |> dplyr::slice(1)
v_total <- tv$vc + tv$vp + tv$vp2

structural <- tibble::tibble(
  Quantity = c("Total apparent volume Vc/F + Vp1/F + Vp2/F (L)",
               "Transit rate constant ktr = 3 / MTT (1/h)",
               "Relative bioavailability F1 (unitless)"),
  Model     = c(v_total, tv$ktr, tv$fdepot),
  Published = c(1206, 3 / 1.78, 1)
)
knitr::kable(structural, digits = 4,
             caption = "Structural identities against Gaffney 2026 (Discussion; Table 3).")

# Deterministic quantities -- exact agreement is the correct expectation.
stopifnot(
  abs(v_total - 1206) < 1e-6,
  abs(tv$ktr - 3 / 1.78) < 1e-9,
  abs(tv$fdepot - 1) < 1e-9
)
```

# PKNCA validation

```{r pknca-scenarios}
scen_conc <- scen_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, scenario)

# Guarantee a time-zero record per subject; pre-dose Cc is 0 for an oral dose.
scen_conc <- dplyr::bind_rows(
  scen_conc,
  scen_conc |> dplyr::distinct(id, scenario) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, scenario, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

scen_dose <- scen_events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, scenario)

scen_intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE,
  aucinf.obs = TRUE, aucpext.obs = TRUE, half.life = TRUE
)

scen_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(scen_conc, Cc ~ time | scenario + id),
  PKNCA::PKNCAdose(scen_dose, amt ~ time | scenario + id),
  intervals = scen_intervals
))

scen_res <- as.data.frame(scen_nca$result) |>
  dplyr::select(scenario, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
```

## AUC(0-inf) reproduces `F1 * Dose / (CL/F)`

For a linear model with complete absorption the extrapolated AUC must equal
`F1 * Dose / (CL/F)` exactly. This is the strongest single test that the transit
chain, the three-compartment disposition and the bioavailability term are all
wired together correctly, because any mis-wiring loses or duplicates dose.

```{r auc-identity}
auc_check <- scen_res |>
  dplyr::select(scenario, aucinf.obs, aucpext.obs) |>
  dplyr::left_join(
    scen_sim |> dplyr::distinct(scenario, .keep_all = TRUE) |>
      dplyr::transmute(scenario, cl, fdepot),
    by = "scenario"
  ) |>
  # `central` is mg and `vc` is L, so Cc is scaled by 1000 to ng/mL; the
  # analytic AUC needs the same factor.
  dplyr::mutate(
    analytic = fdepot * 300 / cl * 1000,
    pct_diff = 100 * (aucinf.obs - analytic) / analytic
  )

auc_check |>
  dplyr::transmute(
    Scenario = scenario,
    `AUC0-inf, NCA` = signif(aucinf.obs, 6),
    `F1 * Dose / CL` = signif(analytic, 6),
    `Difference (%)` = signif(pct_diff, 3),
    `Extrapolated (%)` = signif(aucpext.obs, 3)
  ) |>
  knitr::kable(caption = "NCA AUC(0-inf) against the closed-form oral AUC.")

# Deterministic: this is pure numerical error, so a tight bound is correct.
stopifnot(max(abs(auc_check$pct_diff)) < 0.05)
```

## Published exposure ratios (Figure 5B)

The AUC ratios below are transcribed from the Results text accompanying
Figure 5B. They are the paper's own translation of the covariate coefficients
into exposure, so reproducing all thirteen of them checks every covariate
coefficient, every reference value, and the fed fractional change on F1 at once.

```{r auc-ratios}
published_ratios <- tibble::tribble(
  ~scenario,             ~published, ~source,
  "ALB 1.7 g/dL (min)",  1.89,  "Results: 1.89-fold (90% CI 1.65-2.13)",
  "ALB 3.1 g/dL (P5)",   1.21,  "Results: 1.21 (90% CI 1.16-1.25)",
  "ALB 4.7 g/dL (P95)",  0.89,  "Results: 0.89 (90% CI 0.87-0.91)",
  "ALB 7.9 g/dL (max)",  0.60,  "Results: 0.60-fold (90% CI 0.55-0.67)",
  "CrCL 27.6 (min)",     1.37,  "Results: 1.37-fold (90% CI 1.31-1.43)",
  "CrCL 44.3 (P5)",      1.20,  "Results: 1.20 (95% CI 1.16-1.23)",
  "CrCL 147.4 (P95)",    0.85,  "Results: 0.85 (90% CI 0.83-0.87)",
  "CrCL 298.6 (max)",    0.69,  "Results: 0.69-fold (90% CI 0.66-0.73)",
  "ALP 0.9 (min)",       0.72,  "Results: 0.72-fold (90% CI 0.64-0.79)",
  "ALP 49 (P5)",         0.96,  "Results: 0.96 (90% CI 0.95-0.97)",
  "ALP 211.8 (P95)",     1.07,  "Results: 1.07 (90% CI 1.05-1.09)",
  "ALP 1814 (max)",      1.25,  "Results: 1.25-fold (90% CI 1.17-1.35)",
  "Fed",                 1.236, "Table 3: F1 * (1 + 0.236)"
)

ref_auc <- scen_res$aucinf.obs[scen_res$scenario == "Reference"]
stopifnot(length(ref_auc) == 1L)

ratio_cmp <- published_ratios |>
  dplyr::left_join(scen_res |> dplyr::select(scenario, aucinf.obs), by = "scenario") |>
  dplyr::mutate(
    simulated = aucinf.obs / ref_auc,
    absdiff   = abs(simulated - published)
  )
stopifnot(nrow(ratio_cmp) == 13L, !anyNA(ratio_cmp$simulated))

ratio_cmp |>
  dplyr::transmute(
    Scenario = scenario,
    `Simulated AUC ratio` = round(simulated, 3),
    `Published AUC ratio` = published,
    `Absolute difference` = round(absdiff, 4),
    Source = source
  ) |>
  knitr::kable(caption = paste(
    "Simulated steady-state-equivalent AUC ratios against the values Gaffney",
    "2026 reports for Figure 5B. Published ratios are printed to two decimal",
    "places."
  ))

# Deterministic (typical-value) quantities. The published ratios are rounded to
# two decimals, so 0.005 of the residual is pure print rounding; 0.01 leaves
# headroom over the largest realised difference (0.0064, ALP maximum) while
# still failing on any mis-transcribed exponent or reference value -- a 10%
# error in e_alb_cl alone moves the ALB minimum ratio by more than 0.15.
stopifnot(max(ratio_cmp$absdiff) < 0.01)
```

```{r figure-5b, fig.height = 5}
# Replicates Figure 5B of Gaffney 2026: forest plot of covariate effects on
# steady-state AUC relative to the reference patient.
ratio_cmp |>
  dplyr::mutate(scenario = factor(scenario, levels = rev(published_ratios$scenario))) |>
  tidyr::pivot_longer(c(simulated, published), names_to = "Source", values_to = "ratio") |>
  dplyr::mutate(Source = dplyr::recode(Source,
                                       simulated = "nlmixr2lib simulation",
                                       published = "Gaffney 2026 reported")) |>
  ggplot(aes(ratio, scenario, colour = Source, shape = Source)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey40") +
  geom_point(size = 2.6, alpha = 0.8) +
  scale_x_continuous(breaks = seq(0.6, 2.0, by = 0.2)) +
  labs(x = "AUC ratio vs reference patient", y = NULL,
       title = "Figure 5B -- covariate effects on niraparib AUC",
       caption = "Replicates Figure 5B of Gaffney 2026.") +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Terminal half-life and Tmax

The three-compartment disposition has analytic half-lives of 3.03, 34.0 and
136.0 h (eigenvalues of the rate-constant matrix at the typical values). NCA over
the 1000 h window should recover the terminal phase.

```{r half-life}
micro <- tv
rate_matrix <- matrix(c(
  -(micro$kel + micro$k12 + micro$k13), micro$k21, micro$k31,
  micro$k12,                           -micro$k21, 0,
  micro$k13,                            0,         -micro$k31
), nrow = 3, byrow = TRUE)
analytic_hl <- sort(log(2) / -Re(eigen(rate_matrix)$values))

hl_nca <- scen_res$half.life[scen_res$scenario == "Reference"]
tmax_fasted <- scen_res$tmax[scen_res$scenario == "Reference"]
tmax_fed <- scen_res$tmax[scen_res$scenario == "Fed"]

tibble::tibble(
  Quantity = c("Terminal half-life (h)", "Tmax, fasted/unknown (h)",
               "Tmax, fed (h)"),
  Model = c(hl_nca, tmax_fasted, tmax_fed),
  `Analytic / expected` = c(max(analytic_hl), NA_real_, NA_real_)
) |>
  knitr::kable(digits = 2, caption = "Terminal half-life and Tmax at typical values.")

stopifnot(
  # NCA fits the terminal slope over a finite window, so allow 3%.
  abs(hl_nca - max(analytic_hl)) / max(analytic_hl) < 0.03,
  # Figure 4 shows the TABLET and HEPATIC pcVPC medians peaking within the
  # first few hours after dose; both typical Tmax values must sit in that band.
  tmax_fasted > 1.5, tmax_fasted < 6,
  # A 115% increase in MTT must move Tmax substantially later.
  tmax_fed > tmax_fasted + 1.5
)
```

# Virtual cohort

Individual patient data are not public. The cohort below draws covariates from
log-normal distributions matched to the percentiles the paper reports for
Figure 5 (ALB, ALP, CrCL); body weight uses the 70 kg reference with an assumed
20% coefficient of variation, because the weight distribution lives in
Supplementary Table S1, which is not in the open-access deposit.

```{r cohort}
# set.seed() seeds R's RNG for the covariate draws. It does NOT seed rxode2's
# simulation RNG, whose streams are partitioned per solver thread -- so the etas
# below differ between a 2-core CI runner and a 16-thread workstation. Every
# assertion downstream is written to hold for any cohort the model can produce.
set.seed(20260901)
rxode2::rxSetSeed(20260901)

n_per_arm <- 150

# sigma on the log scale that reproduces the reported 5th and 95th percentiles.
log_sd <- function(p05, p95) (log(p95) - log(p05)) / (2 * stats::qnorm(0.95))

make_arm <- function(n, fed, label, id_offset) {
  tibble::tibble(
    id           = id_offset + seq_len(n),
    arm          = label,
    FED          = fed,
    ALB          = 40    * exp(stats::rnorm(n, 0, log_sd(31, 47))),
    ALP          = 83    * exp(stats::rnorm(n, 0, log_sd(49, 211.8))),
    CRCL         = 82.37 * exp(stats::rnorm(n, 0, log_sd(44.3, 147.4))),
    WT           = 70    * exp(stats::rnorm(n, 0, 0.20)),
    # TABLET is the study that supplied the intensively sampled single-dose
    # data these profiles emulate, so its residual magnitude is selected.
    STUDY_NOVA   = 0, STUDY_QUADRA = 0, STUDY_PRIMA = 0,
    STUDY_TABLET = 1, STUDY_HEPATIC = 0
  )
}

cohort <- dplyr::bind_rows(
  make_arm(n_per_arm, 0, "Fasted/unknown", id_offset = 0L),
  make_arm(n_per_arm, 1, "Fed",            id_offset = 1000L)
)

cohort_events <- expand_events(cohort, seq(0, 48, by = 0.25))
stopifnot(!anyDuplicated(unique(cohort_events[, c("id", "time", "evid")])))
```

```{r cohort-sim}
cohort_sim <- rxode2::rxSolve(mod, events = cohort_events, keep = "arm") |>
  as.data.frame()

stopifnot(max(cohort_sim$Cc, na.rm = TRUE) > 1)

# The residual-error multiplexer must select the TABLET magnitude for every
# record in this cohort. Deterministic given the study indicators.
stopifnot(isTRUE(all.equal(unique(cohort_sim$expSdCc), 0.324)))
```

```{r figure-4, fig.height = 4}
# Comparable to Figure 4 of Gaffney 2026: simulated concentration percentiles
# over the first 48 h after a single 300 mg dose. Gaffney 2026 plots
# prediction-corrected VPCs against observed data that are not public, so this
# reproduces the shape and magnitude of the profile rather than the pcVPC
# itself.
cohort_sim |>
  # Drop only the pre-dose record so the log-scale axis has no zero.
  dplyr::filter(time != 0) |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(
    Q05 = stats::quantile(Cc, 0.05),
    Q50 = stats::median(Cc),
    Q95 = stats::quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~arm) +
  scale_y_log10() +
  labs(x = "Hours after dose", y = "Niraparib concentration (ng/mL)",
       title = "Simulated 5th / 50th / 95th percentiles, single 300 mg dose",
       caption = "Comparable to Figure 4 of Gaffney 2026 (TABLET panel).") +
  theme_bw()
```

## Cohort NCA by prandial state

```{r pknca-cohort}
cohort_conc <- cohort_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

cohort_conc <- dplyr::bind_rows(
  cohort_conc,
  cohort_conc |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

cohort_dose <- cohort_events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, arm)

cohort_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(cohort_conc, Cc ~ time | arm + id),
  PKNCA::PKNCAdose(cohort_dose, amt ~ time | arm + id),
  intervals = data.frame(start = 0, end = 48,
                         cmax = TRUE, tmax = TRUE, auclast = TRUE)
))

cohort_res <- as.data.frame(cohort_nca$result) |>
  dplyr::select(arm, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

cohort_summary <- cohort_res |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    n = dplyr::n(),
    cmax_med = stats::median(cmax),
    tmax_med = stats::median(tmax),
    auc48_med = stats::median(auclast),
    .groups = "drop"
  )

cohort_summary |>
  dplyr::rename(
    "Prandial state"         = arm,
    "N"                      = n,
    "Median Cmax (ng/mL)"    = cmax_med,
    "Median Tmax (h)"        = tmax_med,
    "Median AUC0-48 (`ng*h/mL`)" = auc48_med
  ) |>
  knitr::kable(digits = 1, caption = paste(
    "Simulated single-dose NCA by prandial state (150 subjects per arm).",
    "Gaffney 2026 reports no NCA table, so these are descriptive; the",
    "quantitative gates are the typical-value checks above."
  ))
```

The paper describes the prandial effects on exposure as "relatively minor" while
MTT more than doubles when fed. Both are visible here: median Tmax moves several
hours later, while median 48 h exposure changes modestly.

```{r prandial-assertions}
fed_row <- cohort_summary[cohort_summary$arm == "Fed", ]
fst_row <- cohort_summary[cohort_summary$arm == "Fasted/unknown", ]
stopifnot(nrow(fed_row) == 1L, nrow(fst_row) == 1L)

# A 115% increase in MTT is a large structural effect, not a near-zero one, so
# a median ordering with headroom is safe across cohorts. Realised gaps were
# 3.50 / 3.00 / 2.75 / 2.50 h at 1 / 2 / 4 / 16 solver threads, so 1 h leaves
# 1.5 h of headroom below the smallest observed value.
stopifnot(fed_row$tmax_med > fst_row$tmax_med + 1)

# "Relatively minor" effect on exposure: bound the MAGNITUDE of the median
# AUC0-48 change, not its sign. This is a descriptive cohort check, not the
# gate on the F1 food effect -- that gate is the deterministic "Fed" row of the
# published-ratio table above, where AUC(0-inf) reproduces 1.236 exactly. Over
# a truncated 48 h window the slower fed absorption partly offsets the higher
# F1, so the ratio lands near 1. Realised 1.104 / 1.203 / 1.080 / 1.205 at
# 1 / 2 / 4 / 16 solver threads; the 0.85-1.45 band leaves >0.23 of headroom on
# each side and still goes red on a sign-flipped fed effect on F1 (which would
# put the ratio near 0.68).
auc_ratio <- fed_row$auc48_med / fst_row$auc48_med
stopifnot(auc_ratio > 0.85, auc_ratio < 1.45)
```

## Study-specific residual error

A separate additive-on-log-scale residual error was estimated for each pooled
study. The model selects it from five binary study indicators, with PN001 as the
reference level.

```{r residual-multiplexer}
study_cols <- c("STUDY_NOVA", "STUDY_QUADRA", "STUDY_PRIMA",
                "STUDY_TABLET", "STUDY_HEPATIC")

study_subjects <- tibble::tibble(
  id    = 1:6,
  study = c("PN001", "NOVA", "QUADRA", "PRIMA", "TABLET", "HEPATIC"),
  ALB = 40, ALP = 83, CRCL = 82.37, WT = 70, FED = 0
)
for (j in seq_along(study_cols)) {
  study_subjects[[study_cols[j]]] <- as.numeric(study_subjects$id == j + 1L)
}

study_sim <- rxode2::rxSolve(
  rxode2::zeroRe(mod),
  events = expand_events(study_subjects, seq(0, 24, by = 1)),
  keep = "study", omega = NA
) |>
  as.data.frame()

study_check <- study_sim |>
  dplyr::group_by(study) |>
  dplyr::summarise(model = unique(expSdCc), .groups = "drop") |>
  dplyr::left_join(
    tibble::tibble(
      study = c("PN001", "NOVA", "QUADRA", "PRIMA", "TABLET", "HEPATIC"),
      published = c(0.227, 0.350, 0.381, 0.451, 0.324, 0.146)
    ),
    by = "study"
  )

study_check |>
  dplyr::rename("Study" = study,
                "Model log-scale residual SD" = model,
                "Gaffney 2026 Table 3" = published) |>
  knitr::kable(digits = 3,
               caption = "Study-specific residual-error selection.")

# Deterministic lookup, so exact equality is the correct assertion.
stopifnot(nrow(study_check) == 6L,
          isTRUE(all.equal(study_check$model, study_check$published)))
```

# Assumptions and deviations

- **The transit rate constant is derived as `ktr = 3 / MTT`, not `4 / MTT`.**
  Table 3 reports MTT but no absorption rate constant, and the paper never
  writes the absorption ODEs out. The commonly used Savic (2007)
  parameterisation is quoted as `MTT = (n + 1) / ktr` for `n` transit
  compartments, which would give `ktr = 4 / 1.78 = 2.247 1/h` here. That reading
  was rejected for three reasons. (i) Savic's `n + 1` counts *first-order
  steps*, not boxes -- the dose lands in a dosing compartment and the chain runs
  to a separate absorption compartment, giving one more step than there are
  labelled transit compartments. Figure 3 has no such extra step: the dose
  arrow enters the first Transit box and the third Transit empties directly into
  Central, so the diagram has exactly three `Ka` arrows between the dose record
  and the central compartment. (ii) The mean of a sum of three exponential steps
  at rate `ktr` is `3 / ktr`, so `MTT = 1.78 h` is self-consistent with the
  name only under this reading. (iii) The typical Tmax is 3.45 h under
  `3 / MTT` and 2.70 h under `4 / MTT`; the TABLET and HEPATIC pcVPC panels in
  Figure 4 show the medians peaking around 4-5 h (a mixture of fasted and fed
  records, whose typical Tmax values are 3.45 and 6.40 h under this reading),
  which the `3 / MTT` reading matches better. If the authors did in fact code
  the Savic `(n + 1)` form with three boxes, absorption in this file is 33%
  slower than theirs; exposure (AUC) is unaffected either way, since MTT does
  not enter the AUC.
- **Albumin unit convention.** The paper reports albumin in the US convention
  g/dL and the canonical `ALB` column is SI g/L. The reference is therefore
  written as 40 g/L rather than 4 g/dL. Because albumin enters only as the ratio
  `(ALB / 40)`, this is a pure relabelling with no effect on any coefficient --
  confirmed by the exact reproduction of all four published albumin AUC ratios.
- **Creatinine clearance is not BSA-normalized.** The canonical `CRCL` column is
  documented in mL/min/1.73 m^2, but Gaffney 2026 reports raw creatinine
  clearance in mL/min (reference 82.37 mL/min) and does not name the estimating
  equation. The model's `covariateData` records the raw-mL/min convention. Raw
  mL/min is the convention used by most `CRCL` models already in the library.
- **Q1/F.** Table 3 reports 43.8 L/h and the Discussion quotes 43.5 L/h for the
  same parameter. The Table 3 value is used, being the parameter table.
- **No off-diagonal IIV.** The paper says alternative variance-covariance
  structures were investigated during covariate modelling, but Table 3 reports
  only the five diagonal IIV magnitudes as CV%, with no covariances or
  correlations. The diagonal form is encoded.
- **Residual error is encoded as `lnorm()`.** The Table 3 note states that "an
  additive residual error model on log-transformed concentrations was used",
  which is exactly nlmixr2's log-normal error structure with the SD on the log
  scale. The tabulated values (labelled "Res variability CV") are taken as those
  log-scale SDs.
- **PN001 is the residual-error reference level.** The paper gives six
  co-equal study-specific magnitudes with no reference. Encoding them as five
  binary indicators plus a bare `expSd` requires nominating one; PN001 was
  chosen as the earliest study and the first row of both Table 1 and the Table 3
  residual block. A record with all five indicators at 0 therefore receives the
  PN001 magnitude.
- **Supplementary Table S1 is not on disk.** The EuropePMC supplementary-files
  endpoint returns no deposit for PMC13185089, so the baseline demographic
  table (age, weight and race distributions) could not be read. The population
  metadata records the percentages quoted in the Results text instead, and the
  virtual cohort's body-weight distribution is an assumption (70 kg median,
  20% CV) rather than a transcription. The albumin, alkaline phosphatase and
  creatinine clearance distributions *are* anchored to published percentiles.
- **No published NCA table.** Gaffney 2026 reports no non-compartmental Cmax /
  Tmax / AUC summary, so there is no `ncaComparisonTable()` comparison here.
  The published quantities that exist -- the thirteen exposure ratios of
  Figure 5B, the total apparent volume of 1206 L, and the six residual-error
  magnitudes -- are each gated above instead.
- **Hepatic impairment is not a model covariate.** Eight patients in HEPATIC had
  moderate hepatic impairment, which the paper names as too few to identify an
  effect. `STUDY_HEPATIC` is a study-membership indicator that selects a
  residual-error magnitude, not a hepatic-function covariate.
