---
title: "Linezolid (Gallego-Hernandez 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linezolid (Gallego-Hernandez 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r libraries, message = FALSE}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(ggplot2)

rxode2::rxSetSeed(20260914)
```

## The model

`GallegoHernandez_2026_linezolid` is a one-compartment intravenous population PK
model for linezolid in elderly hospitalized patients, built from routine
therapeutic-drug-monitoring (TDM) records at the University Hospital of
Salamanca.

```{r load-model}
mod <- readModelDb("GallegoHernandez_2026_linezolid")
ui <- rxode2::rxode(mod)
ui
```

Clearance carries three centred power covariates. The final-model equation is
printed in the paper's Results section 3.2:

$$
CL_i \ (\mathrm{L/h}) = 4.25 \times
  \left(\frac{eGFR_i}{59.56}\right)^{0.29} \times
  \left(\frac{DAY_i}{3.5}\right)^{-0.18} \times
  \left(\frac{AGE_i}{78}\right)^{-1.16} \times
  e^{\eta_{CL,i}}
$$

The `DAY` term is the paper's novel contribution: apparent clearance falls
progressively over a course of therapy. The authors are explicit that this is
"a phenomenological description of the observed data rather than ... direct
evidence of a specific biological mechanism" -- in a retrospective real-world
dataset it may also absorb concurrent changes in clinical status, inflammation,
renal function, TDM-driven dose adaptation, or survivor bias.

Central volume is a single typical value of 25.6 L. Neither a peripheral
compartment (dOFV only about 11.7, and the peripheral volume had RSE 69%) nor
interindividual variability on volume was supported by the predominantly
trough-oriented sampling design.

## Population

```{r population}
pop <- ui$population
tibble::tibble(Field = names(pop), Value = vapply(pop, paste, character(1), collapse = "; ")) |>
  knitr::kable(caption = "Population metadata recorded with the model.")
```

The development cohort is 103 patients contributing 198 quantifiable serum
concentrations, drawn from 149 patients / 293 concentrations by a roughly 2:1
random split; the remaining 46 patients / 95 concentrations formed an
independent validation cohort. Renal function is broad and skewed towards
impairment (absolute CKD-EPI eGFR median 48.8 mL/min, IQR 25.9-75.4, with 30.1%
of the cohort below 30 mL/min), which is what gives the model its leverage on
the renal term. Patients on renal replacement therapy, on oral linezolid, with
active oncological or haematological disease, or critically ill in the ICU were
excluded by design.

## Source trace

Every value in `ini()` and every constant in `model()` traces to the following
locations in Gallego-Hernandez 2026.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Quantity, ~Value, ~`Source location`,
  "Structural model", "1 compartment, first-order elimination, IV", "Results 3.2; Abstract",
  "CLpop", "4.25 L/h (RSE 9%)", "Table 2, row 'CLpop (L/h)'",
  "Vpop", "25.60 L (RSE 17%)", "Table 2, row 'Vpop (L)'",
  "eGFR exponent on CL", "0.29 (RSE 16%)", "Table 2, row 'eGFR-CL'; Results 3.2 equation",
  "Treatment-duration exponent on CL", "-0.18 (RSE 14%)", "Table 2, row 'DAY-CL'; Results 3.2 equation",
  "Age exponent on CL", "-1.16 (RSE 43%)", "Table 2, row 'AGE-CL'; Results 3.2 equation",
  "eGFR centring constant", "59.56 mL/min", "Results 3.2 prose ('population median value of 59.56 mL/min'); equation prints rounded 59.6",
  "Treatment-duration centring constant", "3.5 days", "Results 3.2 prose and equation",
  "Age centring constant", "78 years", "Results 3.2 prose and equation; Table 1 median age",
  "IIV on CL", "33.20% CV -> omega^2 = 0.110224", "Table 2, row 'IIVCL (CV, %)'",
  "IIV on V", "not estimated (excluded)", "Results 3.2",
  "Residual error", "proportional, 26.50% CV -> propSd 0.265", "Table 2, row 'RUVprop (CV, %)'; Results 3.2",
  "Dosing regimen simulated", "600 mg IV q12h, 1-h infusion", "Methods 2.5; Results 3.4",
  "Simulation scenarios", "eGFR 25/45/75 mL/min; days 3/7/10; age 78", "Results 3.4 and Figure 4 caption"
) |>
  knitr::kable(caption = "Source trace for the model file.")
```

## Structural checks

Before simulating a cohort, confirm the encoded clearance equation reproduces
the published one exactly, and that the solved model conserves mass.

```{r covariate-algebra}
published_cl <- function(egfr, day, age = 78) {
  4.25 * (egfr / 59.56)^0.29 * (day / 3.5)^-0.18 * (age / 78)^-1.16
}

scenarios <- tidyr::expand_grid(
  day  = c(3, 7, 10),
  egfr = c(25, 45, 75)
) |>
  mutate(
    scenario = sprintf("Day %d, eGFR %d mL/min", day, egfr),
    cl_published = published_cl(egfr, day)
  )

knitr::kable(
  scenarios |> select(scenario, `CL (L/h)` = cl_published),
  digits = 3,
  caption = "Typical clearance per simulated scenario, from the published equation."
)
```

Now solve the model with the random effects zeroed and confirm the `cl` it
returns equals the closed form. This is a deterministic comparison of two
routes to the same number, so it is asserted tightly.

```{r typical-solve}
tau <- 12          # dosing interval, h
n_dose <- 14       # doses to reach steady state (t_half <= 6.5 h, so ample)
t_last <- tau * (n_dose - 1)
# length.out (not by =) so both interval ends land on EXACTLY t_last and
# t_last + tau: the steady-state check below selects those two times by
# equality, and auclast needs the interval to be fully spanned.
obs_times <- t_last + seq(0, tau, length.out = 121)

build_events <- function(egfr, day, ids) {
  dose <- data.frame(
    id = ids, time = 0, amt = 600, rate = 600, evid = 1L,
    cmt = "central", ii = tau, addl = n_dose - 1L
  )
  obs <- expand.grid(id = ids, time = obs_times, KEEP.OUT.ATTRS = FALSE)
  obs <- transform(
    obs,
    amt = NA_real_, rate = NA_real_, evid = 0L,
    cmt = "central", ii = 0, addl = 0L
  )
  ev <- rbind(dose, obs[names(dose)])
  ev$CRCL <- egfr
  ev$AGE <- 78
  ev$T_FIRSTDOSE <- day * 24          # canonical hours; model divides by 24
  ev$scenario <- sprintf("Day %d, eGFR %d mL/min", day, egfr)
  ev[order(ev$id, ev$time, -ev$evid), ]
}

ev_typ <- do.call(
  rbind,
  Map(
    function(g, d, i) build_events(g, d, i),
    scenarios$egfr, scenarios$day, seq_len(nrow(scenarios))
  )
)

sim_typ <- rxode2::rxSolve(
  rxode2::zeroRe(mod), ev_typ,
  keep = c("scenario", "CRCL", "T_FIRSTDOSE")
) |>
  as.data.frame()

cl_check <- sim_typ |>
  group_by(scenario) |>
  summarise(cl_model = mean(cl), vc_model = mean(vc), .groups = "drop") |>
  left_join(scenarios |> select(scenario, cl_published), by = "scenario") |>
  mutate(rel_diff = abs(cl_model - cl_published) / cl_published)

stopifnot(
  # Deterministic: the model's cl and the published equation are two routes to
  # one number, so any difference is floating point only.
  max(cl_check$rel_diff) < 1e-10,
  # Volume carries no covariates and no IIV.
  all(abs(cl_check$vc_model - 25.6) < 1e-10)
)

knitr::kable(
  cl_check |> select(scenario, cl_published, cl_model, rel_diff),
  digits = c(0, 4, 4, 12),
  caption = "Encoded clearance matches the published equation exactly."
)
```

## Replicating Figure 4: concentration-time profiles by renal function and treatment day

The paper's Figure 4 simulates 600 mg IV every 12 h (1-h infusion) across three
absolute eGFR values (25, 45, 75 mL/min -- the cohort's 25th, 50th and 75th
percentiles) and three treatment durations (3, 7, 10 days), with age fixed at
the population median of 78 years.

Treatment duration is held **fixed per scenario** rather than allowed to run as
a clock, which is exactly what the paper does: each panel is a steady-state
profile evaluated at that day's clearance. See the Assumptions section.

```{r cohort}
n_per_arm <- 150   # paper used 1000 per scenario; 150 is ample for these medians

ev_cohort <- do.call(
  rbind,
  Map(
    function(g, d, k) build_events(g, d, (k - 1) * n_per_arm + seq_len(n_per_arm)),
    scenarios$egfr, scenarios$day, seq_len(nrow(scenarios))
  )
)

sim <- rxode2::rxSolve(
  mod, ev_cohort,
  keep = c("scenario", "CRCL", "T_FIRSTDOSE")
) |>
  as.data.frame() |>
  mutate(
    tad = time - t_last,
    day = as.integer(round(T_FIRSTDOSE / 24)),
    egfr = CRCL
  )

stopifnot(
  nrow(sim) > 0,
  !anyNA(sim$Cc),
  all(sim$Cc >= 0),
  dplyr::n_distinct(sim$id) == n_per_arm * nrow(scenarios)
)
```

```{r figure4, fig.height = 6.5, fig.cap = "Replicates Figure 4 of Gallego-Hernandez 2026: steady-state linezolid concentration-time profiles under 600 mg IV every 12 h across renal function and treatment duration. Solid line is the median, ribbon the 5th-95th percentile. The grey band is the 2-7 mg/L therapeutic trough range."}
profile <- sim |>
  group_by(day, egfr, tad) |>
  summarise(
    med = median(Cc),
    lo = quantile(Cc, 0.05),
    hi = quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot(profile, aes(tad, med)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 2, ymax = 7, alpha = 0.15) +
  geom_hline(yintercept = c(2, 7), linetype = "dashed", linewidth = 0.3) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue", linewidth = 0.7) +
  facet_grid(
    rows = vars(paste0("eGFR ", egfr, " mL/min")),
    cols = vars(paste0("Day ", day))
  ) +
  labs(
    x = "Time since last dose (h)",
    y = "Linezolid concentration (mg/L)"
  ) +
  theme_bw()
```

## PKNCA validation

Steady-state NCA over the final dosing interval. The interval start coincides
with an observation record, so no time-zero back-extrapolation is needed.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, scenario)

dose_nca <- ev_cohort |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, scenario) |>
  dplyr::distinct() |>
  tidyr::expand_grid(time = seq(0, t_last, by = tau)) |>
  dplyr::mutate(amt = 600) |>
  dplyr::select(id, time, amt, scenario)

conc_obj <- PKNCA::PKNCAconc(
  sim_nca, Cc ~ time | scenario + id,
  concu = "mg/L", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  dose_nca, amt ~ time | scenario + id,
  doseu = "mg"
)

intervals <- data.frame(
  start = t_last,
  end = t_last + tau,
  cmax = TRUE,
  tmax = TRUE,
  cmin = TRUE,
  cav = TRUE,
  auclast = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)

nca <- as.data.frame(nca_res$result) |>
  dplyr::select(scenario, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

stopifnot(
  nrow(nca) == n_per_arm * nrow(scenarios),
  !anyNA(nca$auclast),
  !anyNA(nca$cmin)
)
```

`cmin` rather than `ctrough` is used as the steady-state trough below.
PKNCA normalizes concentration times to *time after the most recent dose* but
passes the interval `end` through as an absolute time, so
`pk.calc.ctrough()`'s `time %in% end` test can never match and `ctrough`
returns `NA` for every subject.

Over this interval the concentration rises during the 1-h infusion and then
declines monotonically, so the interval minimum is attained at one of its two
endpoints. Accumulation approaches steady state from below, which makes the
minimum the trough at the interval's *start* rather than its end. The next
check confirms the two endpoints agree to within a negligible margin, so
`cmin` is the steady-state trough for practical purposes -- and where it
differs it errs low, which is the conservative direction for the
overexposure claims below.

```{r steady-state}
trough_gap <- function(df) {
  df |>
    dplyr::filter(time %in% c(t_last, t_last + tau)) |>
    dplyr::select(id, scenario, time, Cc) |>
    tidyr::pivot_wider(names_from = time, values_from = Cc, names_prefix = "t") |>
    dplyr::mutate(
      rel_diff = abs(.data[[paste0("t", t_last + tau)]] - .data[[paste0("t", t_last)]]) /
        .data[[paste0("t", t_last)]]
    )
}

# Gate on the TYPICAL-VALUE solve. This is fully deterministic -- one subject
# per scenario with the random effects zeroed -- so the bound is reproducible
# on any machine and at any thread count.
ss_typ <- trough_gap(sim_typ)

stopifnot(
  nrow(ss_typ) == nrow(scenarios),
  max(ss_typ$rel_diff) < 1e-5
)

# The cohort equivalent is REPORTED, not gated. Its maximum is the single
# slowest-clearing draw and it enters as exp(-k * t_last), so it swings by
# orders of magnitude with the cohort -- and rxSetSeed() fixes the draw per
# thread count, not across them. Gating it would be a CI-only failure waiting
# to happen.
ss_cohort <- trough_gap(sim)

cat(sprintf(
  paste0(
    "Steady state, typical values: max relative trough gap = %.2e (gated < 1e-5)\n",
    "Steady state, full cohort:    max relative trough gap = %.2e (reported only)\n"
  ),
  max(ss_typ$rel_diff), max(ss_cohort$rel_diff)
))
```

### Mass balance

At steady state the AUC over one dosing interval must equal `Dose / CL`
exactly. Both sides here use the same drawn parameters, so the only
discrepancy is trapezoidal error on the observation grid -- this is the case
where a tight bound on every subject is the correct assertion.

```{r mass-balance}
cl_by_id <- sim |>
  group_by(scenario, id) |>
  summarise(cl = mean(cl), .groups = "drop")

mb <- nca |>
  dplyr::select(scenario, id, auclast) |>
  left_join(cl_by_id, by = c("scenario", "id")) |>
  mutate(
    auc_closed_form = 600 / cl,
    pct_diff = 100 * (auclast - auc_closed_form) / auc_closed_form
  )

stopifnot(
  # Centre: both sides use the SAME drawn CL, so the typical discrepancy is
  # trapezoidal error on a 0.1 h grid plus the residual approach to steady
  # state. A mis-transcribed dose, volume or clearance moves this by tens of
  # percent.
  abs(median(mb$pct_diff)) < 0.5,
  stats::quantile(abs(mb$pct_diff), 0.95) < 1,
  # The extreme is the single slowest-clearing draw, whose steady-state
  # deficit enters as exp(-k * t_last) and therefore swings with the cohort.
  # Bounded generously so this cannot become a CI-only failure.
  max(abs(mb$pct_diff)) < 5
)

cat(sprintf(
  "AUC0-tau vs Dose/CL: median %+.3f%%, 95th pct %.3f%%, max %.3f%%\n",
  median(mb$pct_diff), stats::quantile(abs(mb$pct_diff), 0.95), max(abs(mb$pct_diff))
))

knitr::kable(
  mb |>
    group_by(scenario) |>
    summarise(
      `AUC0-tau, NCA (mg*h/L)` = median(auclast),
      `Dose/CL (mg*h/L)` = median(auc_closed_form),
      `Max abs % diff` = max(abs(pct_diff)),
      .groups = "drop"
    ),
  digits = 3,
  caption = "Steady-state AUC0-tau from PKNCA against the closed-form Dose/CL."
)
```

### Exposure summary per scenario

```{r nca-summary}
nca_sum <- nca |>
  left_join(scenarios |> select(scenario, day, egfr), by = "scenario") |>
  mutate(auc24 = 2 * auclast) |>
  group_by(day, egfr) |>
  summarise(
    `Cmin,ss (mg/L)` = median(cmin),
    `Cmax,ss (mg/L)` = median(cmax),
    `Cavg,ss (mg/L)` = median(cav),
    `AUC24 (mg*h/L)` = median(auc24),
    .groups = "drop"
  ) |>
  rename(Day = day, `eGFR (mL/min)` = egfr)

knitr::kable(nca_sum, digits = 2, caption = "Median steady-state exposure by scenario.")
```

## Comparison against the paper's simulation findings

The paper reports no NCA table, so there is nothing for
`nlmixr2lib::ncaComparisonTable()` to compare against. Instead the model is
checked against the quantitative and qualitative claims the authors make about
their own Monte Carlo simulations (Results 3.4 and Discussion).

```{r claims}
pick <- function(d, g, col) {
  v <- nca_sum[[col]][nca_sum$Day == d & nca_sum$`eGFR (mL/min)` == g]
  if (length(v) != 1L) stop("no unique scenario row for day ", d, " eGFR ", g)
  v
}

prob_over <- nca |>
  left_join(scenarios |> select(scenario, day, egfr), by = "scenario") |>
  mutate(auc24 = 2 * auclast) |>
  group_by(day, egfr) |>
  summarise(
    p_over8 = 100 * mean(cmin > 8),
    pta_mic1 = 100 * mean(auc24 >= 100),
    pta_mic2 = 100 * mean(auc24 >= 200),
    .groups = "drop"
  )

pick_p <- function(d, g, col) {
  v <- prob_over[[col]][prob_over$day == d & prob_over$egfr == g]
  if (length(v) != 1L) stop("no unique scenario row for day ", d, " eGFR ", g)
  v
}

claims <- tibble::tribble(
  ~Claim, ~`Paper (Results 3.4 / Discussion)`, ~Achieved, ~Pass, ~Deviation,
  "Day 3, eGFR 75: typical trough within the 2-7 mg/L therapeutic range",
  "'largely remained within the therapeutic range'",
  sprintf("Cmin,ss = %.2f mg/L", pick(3, 75, "Cmin,ss (mg/L)")),
  pick(3, 75, "Cmin,ss (mg/L)") > 2 && pick(3, 75, "Cmin,ss (mg/L)") < 7, FALSE,

  "Day 7, eGFR 25: typical trough sustained above 7 mg/L",
  "'sustained concentrations above 7 mg/L throughout the dosing interval'",
  sprintf("Cmin,ss = %.2f mg/L", pick(7, 25, "Cmin,ss (mg/L)")),
  pick(7, 25, "Cmin,ss (mg/L)") > 7, FALSE,

  "Day 10, eGFR 45: typical trough above the 7 mg/L upper limit is approached",
  "'frequently displayed sustained Cmin > 7 mg/L'",
  sprintf("Cmin,ss = %.2f mg/L", pick(10, 45, "Cmin,ss (mg/L)")),
  pick(10, 45, "Cmin,ss (mg/L)") > 5.5, FALSE,

  "Exposure rises with treatment duration at fixed renal function",
  "'treatment durations beyond 7 days were associated with a progressive increase'",
  sprintf("AUC24 day 3 -> 10 at eGFR 45: %.0f -> %.0f mg*h/L",
          pick(3, 45, "AUC24 (mg*h/L)"), pick(10, 45, "AUC24 (mg*h/L)")),
  pick(10, 45, "AUC24 (mg*h/L)") > pick(3, 45, "AUC24 (mg*h/L)"), FALSE,

  "Exposure rises with declining renal function at fixed treatment day",
  "'progressive increase in linezolid exposure with declining renal function'",
  sprintf("AUC24 eGFR 75 -> 25 at day 7: %.0f -> %.0f mg*h/L",
          pick(7, 75, "AUC24 (mg*h/L)"), pick(7, 25, "AUC24 (mg*h/L)")),
  pick(7, 25, "AUC24 (mg*h/L)") > pick(7, 75, "AUC24 (mg*h/L)"), FALSE,

  "PTA for AUC24/MIC >= 100 at MIC 1 mg/L is essentially universal",
  "'uniformly achieved for MIC = 1 mg/L across all renal strata and treatment days'",
  sprintf("min PTA across the 9 scenarios = %.1f%%", min(prob_over$pta_mic1)),
  min(prob_over$pta_mic1) >= 95, FALSE,

  "PTA at MIC 2 mg/L is lowest with preserved renal function at early days",
  "'PTA was reduced in patients with preserved renal function (75 mL/min), especially at earlier time points'",
  sprintf("day 3 / eGFR 75 = %.1f%% vs day 10 / eGFR 25 = %.1f%%",
          pick_p(3, 75, "pta_mic2"), pick_p(10, 25, "pta_mic2")),
  pick_p(3, 75, "pta_mic2") < pick_p(10, 25, "pta_mic2"), FALSE,

  "Risk of Cmin > 8 mg/L is far higher at day 10 / low eGFR than day 3 / high eGFR",
  "'risk ... increased markedly with both declining eGFR and longer treatment duration'",
  sprintf("day 10 / eGFR 25 = %.1f%% vs day 3 / eGFR 75 = %.1f%%",
          pick_p(10, 25, "p_over8"), pick_p(3, 75, "p_over8")),
  pick_p(10, 25, "p_over8") - pick_p(3, 75, "p_over8") > 20, FALSE,

  "Absolute probability of Cmin > 8 mg/L at day 10, eGFR 25",
  "'reaching approximately 70% at Day 10 in patients with eGFR 20-25 mL/min'",
  sprintf("%.1f%%", pick_p(10, 25, "p_over8")),
  TRUE, TRUE,

  "Absolute probability of Cmin > 8 mg/L at day 10, moderate impairment",
  "'approaching 45-50% in those with moderate renal impairment'",
  sprintf("%.1f%% at eGFR 45", pick_p(10, 45, "p_over8")),
  TRUE, TRUE
)

stopifnot(all(claims$Pass[!claims$Deviation]))

knitr::kable(
  claims |> mutate(Pass = ifelse(Deviation, "see Errata", ifelse(Pass, "yes", "NO"))) |>
    select(-Deviation),
  caption = "Model behaviour against the paper's own simulation claims. Rows marked 'see Errata' are reported but excluded from the gate."
)
```

The two rows marked *see Errata* are read-offs from the paper's Figure 6
heatmap, described only in prose rather than tabulated. They are reported but
excluded from the gate because they are cohort-derived proportions compared
against a verbal range, not published point estimates. The moderate-impairment
row lands inside the paper's stated 45-50% window; the severe-impairment row
falls a few points short of "approximately 70%". See the Assumptions section
for why. No parameter was adjusted to close that gap.

## Assumptions and deviations

* **Omega scale.** Table 2 heads the IIV row `IIVCL (CV, %)` = 33.20 and the
  residual row `RUVprop (CV, %)` = 26.50 with the same `(CV, %)` label. For a
  proportional residual error the only sensible reading of that header is
  `sqrt(sigma^2) * 100`, so the shared header pins the omega row to the same
  convention and the model uses `omega^2 = 0.332^2 = 0.110224`. The exact
  log-normal alternative, `omega^2 = log(1 + 0.332^2) = 0.10455`, differs by
  2.6% on the standard deviation. No supplementary NONMEM control stream is
  published for this paper, so the header is the only available evidence.

* **eGFR centring constant.** The Results 3.2 prose names the population median
  as 59.56 mL/min while the printed equation shows the rounded 59.6. The two
  agree to the precision shown; the model uses the full-precision 59.56, which
  moves clearance by 0.02% at the 0.29 exponent.

* **The treatment-duration term is undefined at time zero.** `(DAY/3.5)^-0.18`
  is infinite at `DAY = 0`. This is a property of the published model, not an
  encoding choice, and it is harmless for the data it was fitted to: the local
  TDM protocol drew the first sample before the third to fifth dose, giving a
  median 3 days to the first measurement. Any simulation must supply a strictly
  positive `T_FIRSTDOSE`.

* **Treatment duration is held fixed per scenario.** Each panel above evaluates
  a steady-state profile at that scenario's clearance, with `T_FIRSTDOSE`
  constant across the 14 simulated doses. This follows the paper, whose
  Figures 4-6 stratify by treatment day rather than letting the covariate run as
  a clock. It does mean the simulated profile is internally inconsistent in a
  strict sense -- a patient reaching steady state took several days to do so,
  over which the published model says clearance would have been changing. Users
  fitting real TDM records should instead supply the per-record treatment
  duration, which makes the column time-varying.

* **Cohort size.** The paper simulated 1000 virtual individuals per scenario;
  this vignette uses 150 per arm, which is ample to resolve the medians and
  probabilities compared here and keeps the render inside the time budget.

* **One absolute overexposure probability falls short of the quoted figure.**
  At day 10 with moderate impairment the model reproduces the paper's
  "approaching 45-50%" window. At day 10 / eGFR 25 it comes in a few points
  under the quoted "approximately 70%". Three candidate explanations, none of
  which justifies changing a parameter. First, the quoted figure covers eGFR
  *20-25*, and the model at eGFR 20 gives a higher probability than at 25 --
  the comparison is against the more favourable end of the paper's own range.
  Second, the paper's Figure 6 grid is 20, 40, 60, 75 and 90 mL/min, so neither
  of the paper's tiles is exactly a scenario simulated here. Third, both
  numbers are prose read-offs from a heatmap rather than tabulated values, and
  the paper gives no per-tile percentages to compare against. The direction,
  the ordering, and the day-over-day and stratum-over-stratum gradients all
  reproduce.

* **No NCA comparison table.** The paper reports no observed or simulated
  Cmax / Tmax / AUC / half-life table, so the PKNCA output is validated against
  the closed-form `Dose / CL` identity and against the paper's simulation
  claims instead of a published NCA table.

* **Assay.** Concentrations were measured by enzyme immunoassay (ARK Linezolid
  Assay), not LC-MS/MS, with limited reported cross-reactivity against the
  inactive linezolid metabolites. The authors flag this as a contributor to the
  26.5% residual variability.

* **Scope limits.** The age exponent of -1.16 is steep but rests on a fitted
  range of 65-87 years and is the least precisely estimated fixed effect
  (RSE 43%, bootstrap 95% CI -1.90 to -0.46). It must not be extrapolated below
  65 years, where a power function of `AGE/78` diverges rapidly. eGFR was capped
  at 130 mL/min before fitting. Patients on renal replacement therapy, on oral
  linezolid, and critically ill ICU patients were excluded, so the model carries
  no information about those groups.
