---
title: "Epetraborole (Ganesan 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Epetraborole (Ganesan 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Ganesan_2022_epetraborole"))$reference`
* Description: `r rxode2::rxode(readModelDb("Ganesan_2022_epetraborole"))$description`
* Article: [Open Forum Infect Dis 2022;9(Suppl 2):ofac492.645](https://doi.org/10.1093/ofid/ofac492.645)
  (IDWeek 2022 poster 593; open access abstract, PMC9752197)

The publication is a conference abstract. Its Table 2 (an image in the
abstract, reprinted unchanged on the poster) gives every structural, IIV, IOV
and residual-error estimate. The abstract and poster do not print the number
of transit compartments, the allometric exponents, the scale of the IIV on
bioavailability, or whether the ELF compartment depletes the central
compartment. Those four choices are set out, with the evidence used to make
each one, under "Structural choices not printed in the source".

The validation below uses published results that were not used to build this
file:

* Eckburg et al., IDWeek 2022 poster 1727 (Open Forum Infect Dis
  2022;9(Suppl 2):ofac492.1357), Table 2: non-compartmental Day 1 and Day 28
  plasma PK of the phase 1b tablet study (NCT04892641). Its first four
  cohorts were part of the model-building data.
* Tenero D et al. Antimicrob Agents Chemother 2013;57:3334-3339
  ([doi:10.1128/AAC.02483-12](https://doi.org/10.1128/AAC.02483-12)): plasma
  and ELF PK after 1500 mg IV (the "Phase 1, ELF study" of the model-building
  data).
* Patel G et al., IDWeek 2023 poster 2144 (Open Forum Infect Dis
  2023;10(Suppl 2):ofad500.1767): single 500 mg oral dose in adults with
  normal renal function (not part of the model-building data).

# Population

The pooled dataset held 2637 epetraborole PK samples from 138 subjects in
seven studies (abstract Table 1): a first-in-human single/multiple ascending
IV dose study (200-3000 mg single, 500-1600 mg BID), an IV ELF study
(1500 mg single dose and BID), an IV mass-balance study (1500 mg), an oral
single/multiple ascending dose and food-effect study (500-4000 mg single,
2000-3000 mg BID for 10 days), two phase 2 IV studies in complicated urinary
tract infection / acute pyelonephritis and complicated intra-abdominal
infection (750 or 1500 mg BID), and the first four cohorts of an oral tablet
phase 1b study (250, 500, 750 mg QD and 500 mg QOD for 28 days). No patients
with Mycobacterium avium complex lung disease contributed data; body weight
was modelled allometrically so that the lower body weight of that population
could be simulated. Age, weight, sex and race distributions are not reported.

The same information is available programmatically via
`readModelDb("Ganesan_2022_epetraborole")()$population`.

# Source trace

Every `ini()` value carries an in-file comment pointing at the source. "T2"
is Table 2 of the abstract (identical to poster Table 2).

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL, L/h per 70 kg) | log(15.3) | T2 'CL (L/h/70 kg)' |
| `lq` (CLd, L/h per 70 kg) | log(23) | T2 'Cld (L/h/70 kg)' |
| `lq2` (CLd2, L/h per 70 kg) | log(43.3) | T2 'Cld2 (L/h/70 kg)' |
| `lvc` (Vc, L per 70 kg) | log(15.6) | T2 'Vc (L/70 kg)' |
| `lvp` (Vp, L per 70 kg) | log(140) | T2 'Vp (L/70 kg)' |
| `lvp2` (Vp2, L per 70 kg) | log(33.2) | T2 'Vp2 (L/70 kg)' |
| `e_wt_cl`, `e_wt_vc` | 0.75, 1 (fixed) | Not printed; Results 'allometric scaling approach' (see below) |
| `lka_fasted` (1/h) | log(2.89) | T2 'Ka, fast' |
| `lka_fed` (1/h) | log(1.34) | T2 'Ka, fed' |
| `logitfdepot` | logit(0.565) | T2 'F' |
| `lk_central_elf` (K14, 1/h) | log(0.266) | T2 'K14' |
| `lkeff_elf` (K40, 1/h) | log(0.661) | T2 'K40' |
| `etalcl`, `etalvc`, `etalvp`, `etalvp2` | 0.079^2, 0.372^2, 0.0868^2, 0.189^2 | T2 IIV column 7.9, 37.2, 8.68, 18.9 |
| `etalka` | 0.295^2 | T2 IIV on 'Ka, fast' 29.5; shared by the fed ka |
| `etalogitfdepot` | 0.327^2 (logit scale) | T2 IIV on F 32.7 |
| `etalk_central_elf` | 0.578^2 | T2 IIV on K14 57.80 |
| `etaiov_ka_1`, `etaiov_ka_2` | 0.0343 (shared) | T2 'IOV in Ka' 0.0343 (18.5), shrinkage for occasions 1 and 2 |
| `propSd` | sqrt(0.0587) | T2 'RVplasma' 0.0587 (24.2%) |
| `propSd_Celf` | sqrt(0.0656) | T2 'RVELF' 0.0656 (25.6%) |
| Three-compartment disposition, linear elimination | n/a | Results; T2 footnote |
| depot -> transit1 -> transit2 -> central at rate ka | n/a | Results 'transit compartments with separate absorption rates for fed and fasted' (count: see below) |
| `d/dt(elf) <- K14 * central - K40 * elf`, central depleted by K14 | n/a | T2 footnote 'K14, rate constant from central compartment to ELF compartment; K40, rate constant out of the ELF compartment' |
| `Celf <- elf / vc` | n/a | Not printed; see below |

The IIV and IOV variances follow from the table's own convention: the IOV
row prints the estimate 0.0343 next to a magnitude of 18.5, which is
`100 * sqrt(0.0343)`, and the residual rows pair 0.0587 with 24.2%
(`100 * sqrt(0.0587)`). Every magnitude in the IIV column is therefore read
as `100 * sqrt(omega^2)`.

# Structural choices not printed in the source

```{r load-model}
mod <- readModelDb("Ganesan_2022_epetraborole")
ui <- rxode2::rxode(mod)
th <- ui$theta
om <- ui$omega
```

## Number of transit compartments

The Results say oral absorption used "transit compartments with separate
absorption rates for fed and fasted administration", and Table 2 lists only
`Ka, fast` and `Ka, fed`. With no separate transit rate in the table, the
transit rate equals ka. The number of transit compartments is not printed.
The phase 1b study reports a median Tmax of 1.0 h fasted and 2.0 h (range
2.0-3.0) fed after 500 mg (Eckburg 2022 Table 2, cohort 7). The typical-value
Tmax for each candidate chain length:

```{r transit-count}
# Typical subject (70 kg); chain of n transit compartments between depot and
# central, every step at ka. Disposition values are taken from the packaged
# model so only the chain length varies.
chain_model <- function(n) {
  states <- c("depot", if (n > 0) paste0("tr", seq_len(n)))
  odes <- "d/dt(depot) <- -ka * depot"
  for (i in seq_along(states)[-1]) {
    odes <- c(odes, sprintf("d/dt(%s) <- ka * %s - ka * %s", states[i], states[i - 1], states[i]))
  }
  rxode2::rxode2(paste(c(
    sprintf("cl <- %.10g; q <- %.10g; q2 <- %.10g; vc <- %.10g; vp <- %.10g; vp2 <- %.10g; k14 <- %.10g",
            exp(th[["lcl"]]), exp(th[["lq"]]), exp(th[["lq2"]]), exp(th[["lvc"]]),
            exp(th[["lvp"]]), exp(th[["lvp2"]]), exp(th[["lk_central_elf"]])),
    odes,
    sprintf("d/dt(central) <- ka * %s - (cl + q + q2) / vc * central - k14 * central + q / vp * p1 + q2 / vp2 * p2",
            states[length(states)]),
    "d/dt(p1) <- q / vc * central - q / vp * p1",
    "d/dt(p2) <- q2 / vc * central - q2 / vp2 * p2",
    "Cc <- central / vc"
  ), collapse = "\n"))
}
ka_values <- c(fasted = exp(th[["lka_fasted"]]), fed = exp(th[["lka_fed"]]))
grid <- seq(0, 12, by = 0.01)
tmax_tab <- expand.grid(n_transit = 0:4, state = names(ka_values), stringsAsFactors = FALSE)
tmax_tab$tmax <- mapply(function(n, st) {
  s <- as.data.frame(rxode2::rxSolve(chain_model(n),
    rxode2::et(amt = 500, cmt = "depot") |> rxode2::et(grid),
    params = c(ka = ka_values[[st]])))
  s$time[which.max(s$Cc)]
}, tmax_tab$n_transit, tmax_tab$state)
tmax_wide <- tmax_tab |>
  tidyr::pivot_wider(names_from = state, values_from = tmax)
tmax_wide |>
  dplyr::rename(
    "Transit compartments" = n_transit,
    "Typical Tmax fasted (h)" = fasted,
    "Typical Tmax fed (h)" = fed
  ) |>
  knitr::kable(caption = "Observed phase 1b medians: 1.0 h fasted, 2.0 h (range 2.0-3.0) fed.")
```

Only two transit compartments put both the fasted (about 1.1 h) and the fed
(about 2.3 h) typical Tmax at the observed medians; one compartment gives a
fed Tmax below the observed range and three give a fasted Tmax 50% late and a
fed Tmax beyond the observed range. The packaged model therefore uses two
transit compartments.

```{r transit-gate}
# Deterministic typical-value solves: the chosen chain must reproduce both
# observed medians, and its neighbours must not.
two <- tmax_wide[tmax_wide$n_transit == 2, ]
stopifnot(
  abs(two$fasted - 1.0) <= 0.25,
  two$fed >= 2.0, two$fed <= 3.0,
  tmax_wide$fed[tmax_wide$n_transit == 1] < 2.0,
  tmax_wide$fasted[tmax_wide$n_transit == 3] > 1.25
)
```

## Does the ELF compartment deplete plasma?

`K14` and `K40` are NONMEM general-linear (ADVAN5/ADVAN7) microconstant
names: `K14` moves drug from compartment 1 (central) to compartment 4 (ELF)
and `K40` moves it from compartment 4 to the output. Read that way, the ELF
compartment removes 0.266 x 15.6 = 4.15 L/h of plasma clearance-equivalent in
a 70 kg adult, on top of CL. The alternative reading, common for ELF models,
is a link compartment that does not deplete plasma (as in
`Zhang_2021_tazobactam`). Both readings give the same ELF:plasma ratio
(K14 / K40 = 0.40), so the ELF data cannot separate them, but they give
different plasma clearance. Tenero 2013 measured plasma clearance directly
after 1500 mg IV in the ELF study that is part of this dataset: geometric
mean 23.1 L/h after one dose and 20.6 L/h at steady state, half-life 10.7 h,
in 30 mostly male adults with a mean BMI of 24-27 kg/m^2.

```{r elf-drain}
link_mod <- ui |>
  rxode2::model(d/dt(central) <- ka * transit2 - (kel + k12 + k13) * central +
                  k21 * peripheral1 + k31 * peripheral2)

iv_typical <- function(m, wt) {
  # Two endpoints (Cc, Celf): observation rows carry dvid = 1 and no cmt,
  # which returns both concentrations on every row.
  obs_t <- seq(0, 240, by = 0.05)
  ev <- data.frame(
    id = 1L, time = c(0, obs_t), amt = c(1500, rep(NA, length(obs_t))),
    dur = c(1, rep(NA, length(obs_t))), evid = c(1L, rep(0L, length(obs_t))),
    cmt = c("central", rep(NA, length(obs_t))), dvid = c(NA, rep(1L, length(obs_t))),
    WT = wt, FED = 0L, OCC = 1L
  )
  s <- as.data.frame(rxode2::rxSolve(rxode2::zeroRe(m), ev))
  auc <- sum(diff(s$time) * (head(s$Cc, -1) + tail(s$Cc, -1)) / 2)
  term <- s[s$time >= 24 & s$time <= 72, ]
  data.frame(WT = wt, CL = 1500 / auc,
             t_half = -log(2) / unname(coef(lm(log(Cc) ~ time, term))[2]))
}
drain_tab <- dplyr::bind_rows(
  dplyr::bind_rows(lapply(c(70, 80), iv_typical, m = ui)) |> dplyr::mutate(reading = "ELF depletes central (packaged)"),
  dplyr::bind_rows(lapply(c(70, 80), iv_typical, m = link_mod)) |> dplyr::mutate(reading = "ELF link, no depletion")
)
drain_tab |>
  dplyr::select(reading, WT, CL, t_half) |>
  dplyr::rename(
    "ELF reading" = reading, "Weight (kg)" = WT,
    "Typical IV CL (L/h)" = CL, "Half-life 24-72 h (h)" = t_half
  ) |>
  knitr::kable(digits = 1, caption = "Tenero 2013: CL 23.1 L/h (single dose), 20.6 L/h (steady state); half-life 10.7 h.")
```

The depleting reading gives a typical clearance of 19-22 L/h and a half-life
of about 10.5 h, matching Tenero 2013. The link reading gives 15-17 L/h and
12-12.5 h; reaching 23 L/h would need a body weight above 110 kg. Since five
of the seven studies in the model-building data are IV, the fitted model has
to reproduce the IV clearance, so the packaged model lets the ELF
compartment deplete the central compartment. The ELF concentration is the ELF
amount divided by Vc, the only scaling that needs no unprinted volume; it
gives an ELF:plasma AUC ratio of K14 / K40 = 0.40 for a typical subject,
against Tenero's observed 0.54 (see the IV section below).

```{r drain-gate}
# Deterministic typical-value solves at 80 kg (Tenero's subjects: BMI 24-27,
# 93% male). A 15% band around the 21.9 L/h mean of the two Tenero values
# holds the depleting reading (21.7) and excludes the link reading (16.9).
cl80 <- drain_tab$CL[drain_tab$WT == 80]
stopifnot(
  abs(cl80[1] / 21.9 - 1) < 0.15,
  abs(cl80[2] / 21.9 - 1) > 0.15
)
```

## Scale of the IIV on bioavailability

Table 2 gives 32.7% IIV on F = 0.565. On a log scale that would push 4% of
subjects above F = 1 and make oral AUC about 37% variable; on a logit scale F
stays in (0, 1) and oral AUC is about 21% variable. The Results call the
oral variability "slightly higher" than IV, where IIV in CL is 7.9%. The
phase 1b study reports AUC as mean and SD for cohorts of 5-8 subjects; the
coefficients of variation across its 12 Day 1 and Day 28 AUC0-24 entries
have a median of 17%.

```{r f-scale}
set.seed(20221020)
n_draw <- 1e5
wt <- pmin(pmax(rnorm(n_draw, 75, 12), 45), 100)
eta_cl <- rnorm(n_draw, 0, sqrt(om["etalcl", "etalcl"]))
eta_f <- rnorm(n_draw, 0, sqrt(om["etalogitfdepot", "etalogitfdepot"]))
cl <- exp(th[["lcl"]] + eta_cl) * (wt / 70)^0.75
f_logit <- plogis(th[["logitfdepot"]] + eta_f)
f_log <- plogis(th[["logitfdepot"]]) * exp(eta_f)
cohort_cv <- function(auc) {
  replicate(4000, {
    x <- sample(auc, 6)
    100 * sd(x) / mean(x)
  })
}
cv_tab <- data.frame(
  scale = c("logit (packaged)", "log"),
  median_cohort_cv = c(median(cohort_cv(f_logit * 500 / cl)), median(cohort_cv(f_log * 500 / cl))),
  p_f_above_1 = c(mean(f_logit > 1), mean(f_log > 1))
)
cv_tab |>
  dplyr::rename(
    "IIV scale on F" = scale,
    "Median CV of AUC in cohorts of 6 (%)" = median_cohort_cv,
    "Fraction with F > 1" = p_f_above_1
  ) |>
  knitr::kable(digits = 3, caption = "Phase 1b observed: median CV 17% across 12 cohort-day AUC0-24 entries (range 11-29%).")
```

The logit scale reproduces the observed spread; the log scale predicts
cohort CVs near twice the observed values, and under it the observed median
of 17% would be very unlikely. The packaged model therefore puts the IIV on
the logit of F (`logitfdepot`). This is a draw from R's random number
generator only, so the comparison is identical on every machine.

```{r f-scale-gate}
stopifnot(
  abs(cv_tab$median_cohort_cv[1] - 17) < 6,
  cv_tab$median_cohort_cv[2] > 28
)
```

# Virtual cohorts

```{r cohort}
rxode2::rxSetSeed(20221020)
set.seed(20221020)

n_arm <- 100L

# Phase 1b tablets: healthy adults, 40-100 kg (eligibility), weight
# distribution not reported. Occasion 1 is the first dosing day; occasion 2
# every later day.
make_po <- function(n, dose, tau, n_dose, fed, regimen, id_offset) {
  subj <- tibble::tibble(
    id = id_offset + seq_len(n),
    WT = pmin(pmax(rnorm(n, 75, 12), 45), 100),
    FED = fed,
    regimen = regimen
  )
  dose_times <- seq(0, by = tau, length.out = n_dose)
  last <- max(dose_times)
  obs_times <- sort(unique(c(
    seq(0, 24, by = 0.25),
    if (n_dose > 1) last + seq(0, 72, by = 0.25)
  )))
  doses <- tidyr::crossing(subj, time = dose_times) |>
    dplyr::mutate(amt = dose, evid = 1L, cmt = "depot", dvid = NA_integer_)
  obs <- tidyr::crossing(subj, time = obs_times) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = NA_character_, dvid = 1L)
  dplyr::bind_rows(doses, obs) |>
    dplyr::mutate(OCC = ifelse(time < 24, 1L, 2L)) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events_po <- dplyr::bind_rows(
  make_po(n_arm, 250, 24, 28, 0L, "250 mg q24h", 0L),
  make_po(n_arm, 500, 24, 28, 0L, "500 mg q24h", 1000L),
  make_po(n_arm, 750, 24, 28, 0L, "750 mg q24h", 2000L),
  make_po(n_arm, 500, 24, 1, 1L, "500 mg fed", 3000L)
)
stopifnot(!anyDuplicated(unique(events_po[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
sim_po <- rxode2::rxSolve(mod, events = events_po, keep = c("regimen", "WT"),
                          returnType = "data.frame")
stopifnot(!anyNA(sim_po$Cc))
```

## Steady-state profile (poster Figure 3 / abstract Figure 1)

```{r figure-1}
# Replicates the layout of abstract Figure 1 (poster Figure 3): the
# prediction-corrected VPC of phase 1b oral tablet data against time since
# last dose. The observed data are not available, so only the simulated
# 5th, 50th and 95th percentiles are drawn, here for the Day 28 dose.
sim_po |>
  dplyr::filter(regimen != "500 mg fed", time >= 27 * 24) |>
  dplyr::mutate(tad = time - 27 * 24, Cc_ngml = 1000 * Cc) |>
  dplyr::group_by(regimen, tad) |>
  dplyr::summarise(
    Q05 = quantile(Cc_ngml, 0.05), Q50 = median(Cc_ngml), Q95 = quantile(Cc_ngml, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, Q50, colour = regimen, fill = regimen)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.15, colour = NA) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Time since last dose (h)", y = "Epetraborole plasma concentration (ng/mL)",
       colour = NULL, fill = NULL,
       caption = "Median and 90% interval, Day 28 dose. Layout of Ganesan 2022 Figure 1.")
```

# PKNCA validation against the phase 1b study

```{r pknca-po}
conc_po <- sim_po |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, regimen)
dose_po <- events_po |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, regimen)

nca_day <- function(start, end, regimens, half_life) {
  cc <- conc_po |> dplyr::filter(regimen %in% regimens, time >= start, time <= end)
  dd <- dose_po |> dplyr::filter(regimen %in% regimens, time <= start)
  intervals <- data.frame(start = start, end = start + 24, auclast = TRUE, cmax = TRUE, tmax = TRUE)
  if (half_life) {
    intervals <- dplyr::bind_rows(intervals, data.frame(start = start, end = end, half.life = TRUE))
  }
  # bind_rows leaves NA in the logical columns; PKNCA requires TRUE/FALSE.
  flag_cols <- setdiff(names(intervals), c("start", "end"))
  intervals[flag_cols] <- lapply(intervals[flag_cols], function(x) !is.na(x) & x)
  PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(cc, Cc ~ time | regimen + id),
    PKNCA::PKNCAdose(dd, amt ~ time | regimen + id),
    intervals = intervals
  ))
}
nca_d1 <- nca_day(0, 24, unique(events_po$regimen), half_life = FALSE)
nca_d28 <- nca_day(27 * 24, 27 * 24 + 72, c("250 mg q24h", "500 mg q24h", "750 mg q24h"), half_life = TRUE)
```

Eckburg 2022 Table 2 reports arithmetic means for AUC0-24, Cmax and
half-life and medians for Tmax, in ng/mL; they are converted to mg/L here.
The "500 mg q24h" Day 1 row is cohort 3 (n = 6), and "500 mg fed" is the fed
arm of cohort 7.

```{r nca-compare-d1}
published_d1 <- tibble::tribble(
  ~regimen,      ~auclast, ~cmax, ~tmax,
  "250 mg q24h", 6.95,     1.44,  1.03,
  "500 mg q24h", 12.3,     2.62,  1.0,
  "750 mg q24h", 20.0,     5.60,  1.0,
  "500 mg fed",  11.2,     1.75,  2.0
)
cmp_d1 <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_d1, reference = published_d1, by = "regimen",
  units = c(auclast = "mg*h/L", cmax = "mg/L", tmax = "h"),
  tolerance_pct = 20
)
knitr::kable(cmp_d1, caption = "Day 1, simulated vs. Eckburg 2022 Table 2. * differs by more than 20%.")
```

```{r nca-compare-d28}
published_d28 <- tibble::tribble(
  ~regimen,      ~auclast, ~cmax, ~tmax, ~half.life,
  "250 mg q24h", 7.10,     1.39,  1.5,   10.6,
  "500 mg q24h", 13.7,     2.85,  1.0,   10.2,
  "750 mg q24h", 24.4,     5.28,  1.0,   10.3
)
cmp_d28 <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_d28, reference = published_d28, by = "regimen",
  units = c(auclast = "mg*h/L", cmax = "mg/L", tmax = "h", half.life = "h"),
  tolerance_pct = 20
)
knitr::kable(cmp_d28, caption = "Day 28, simulated vs. Eckburg 2022 Table 2. * differs by more than 20%.")
```

```{r nca-gate}
# Robust central tendency only (CLAUDE.md, cohort assertions): a
# mis-transcribed CL, F, V or dose moves these medians by tens of percent.
res_d1 <- as.data.frame(nca_d1)
med <- function(res, reg, param, start) {
  x <- res$PPORRES[res$regimen == reg & res$PPTESTCD == param & res$start == start]
  stopifnot(length(x) > 0)
  median(x, na.rm = TRUE)
}
res_d28 <- as.data.frame(nca_d28)
stopifnot(
  abs(med(res_d1, "500 mg q24h", "auclast", 0) / 12.3 - 1) < 0.3,
  abs(med(res_d1, "250 mg q24h", "auclast", 0) / 6.95 - 1) < 0.3,
  abs(med(res_d28, "500 mg q24h", "auclast", 648) / 13.7 - 1) < 0.3,
  abs(med(res_d1, "500 mg q24h", "tmax", 0) - 1.0) <= 0.5,
  abs(med(res_d1, "500 mg fed", "tmax", 0) - 2.0) <= 0.75
)
```

At 250 and 500 mg the simulated Day 1 and Day 28 AUC0-24, Cmax and
half-life are within about 20% of the phase 1b means, and Tmax reproduces
both the fasted (1.0 h) and fed (2.0 h) medians. The 250 mg Day 28 Tmax
median of 1.5 h against 1.0 h simulated is one sampling step on a cohort of
five. Two departures follow from the model structure:

- The model is linear in dose, but the observed 750 mg cohort (n = 6) had a
  Cmax of 5.6 mg/L on Day 1, 1.4 times the dose-proportional value from the
  500 mg cohort (2.62 mg/L); the simulated 750 mg Cmax and AUC are therefore
  20-29% low. The observed 500 mg Cmax itself ranges from 2.6 to 3.8 mg/L
  across cohorts 2, 3 and 7.
- Food changes only ka in the model, so the simulated fed AUC equals the
  fasted AUC. In the phase 1b food-effect cohort the fed AUC0-24 was 26%
  below the fasted arm (11.2 vs 15.2 mg*h/L); the fed row agrees with the
  reference only because the comparison fasted value (cohort 3, 12.3) is
  lower than cohort 7's.

# IV plasma and ELF exposure (Tenero 2013)

```{r iv-cohort}
n_iv <- 100L
iv_subj <- tibble::tibble(id = seq_len(n_iv), WT = pmin(pmax(rnorm(n_iv, 80, 10), 55), 110))
events_iv <- dplyr::bind_rows(
  iv_subj |> dplyr::mutate(time = 0, amt = 1500, dur = 1, evid = 1L, cmt = "central", dvid = NA_integer_),
  tidyr::crossing(iv_subj, time = c(seq(0, 12, by = 0.25), seq(12.5, 96, by = 0.5))) |>
    dplyr::mutate(amt = NA_real_, dur = NA_real_, evid = 0L, cmt = NA_character_, dvid = 1L)
) |>
  dplyr::mutate(FED = 0L, OCC = 1L, regimen = "1500 mg IV, 1 h") |>
  dplyr::arrange(id, time, dplyr::desc(evid))
sim_iv <- rxode2::rxSolve(mod, events = events_iv, keep = c("regimen", "WT"),
                          returnType = "data.frame")
stopifnot(!anyNA(sim_iv$Cc))

nca_iv <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(sim_iv |> dplyr::select(id, time, Cc, regimen), Cc ~ time | regimen + id),
  PKNCA::PKNCAdose(events_iv |> dplyr::filter(evid == 1) |> dplyr::select(id, time, amt, dur, regimen),
                   amt ~ time | regimen + id, duration = "dur", route = "intravascular"),
  intervals = data.frame(start = 0, end = Inf, cmax = TRUE, aucinf.obs = TRUE,
                         cl.obs = TRUE, vss.obs = TRUE, half.life = TRUE)
))
published_iv <- tibble::tribble(
  ~regimen,          ~cmax, ~aucinf.obs, ~cl.obs, ~vss.obs, ~half.life,
  "1500 mg IV, 1 h", 21.6,  65.0,        23.1,    231,      10.7
)
cmp_iv <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_iv, reference = published_iv, by = "regimen",
  units = c(cmax = "mg/L", aucinf.obs = "mg*h/L", cl.obs = "L/h", vss.obs = "L", half.life = "h"),
  tolerance_pct = 20
)
knitr::kable(cmp_iv, caption = "Single 1500 mg IV dose, simulated vs. Tenero 2013 (geometric means; half-life median). * differs by more than 20%.")
```

```{r elf-ratio}
auc_trap <- function(t, y) sum(diff(t) * (head(y, -1) + tail(y, -1)) / 2)
elf_ratio <- sim_iv |>
  dplyr::group_by(id) |>
  dplyr::summarise(ratio = auc_trap(time, Celf) / auc_trap(time, Cc), .groups = "drop")
# Penetration ratio of the mean profiles, the way Tenero computed it from one
# bronchoscopy per subject at 2, 6 or 12 h.
mean_prof <- sim_iv |>
  dplyr::group_by(time) |>
  dplyr::summarise(Cc = mean(Cc), Celf = mean(Celf), .groups = "drop")
elf_summary <- data.frame(
  quantity = c("Median individual ELF:plasma AUC ratio", "Ratio of mean-profile AUCs", "Tenero 2013 (AUC-based)"),
  value = c(median(elf_ratio$ratio), auc_trap(mean_prof$time, mean_prof$Celf) / auc_trap(mean_prof$time, mean_prof$Cc), 0.538)
)
elf_summary |>
  dplyr::rename("Quantity" = quantity, "ELF:plasma ratio" = value) |>
  knitr::kable(digits = 2)

iv_res <- as.data.frame(nca_iv)
stopifnot(
  abs(median(iv_res$PPORRES[iv_res$PPTESTCD == "cl.obs"]) / 21.9 - 1) < 0.25,
  abs(median(elf_ratio$ratio) - 0.40) < 0.1
)
```

The simulated IV Cmax, AUC, clearance, Vss and half-life are all within
10% of Tenero 2013 (the comparison table labels clearance and Vss "/F",
its generic header; for an IV dose F = 1). The simulated ELF:plasma AUC ratio
(about 0.4, individual or mean-profile) is a quarter below Tenero's 0.54,
which was computed from one bronchoscopy per subject at 2, 6 or 12 h.

# Assumptions and deviations

- **Transit compartments.** Not printed; two compartments, each at the fed or
  fasted ka, chosen because they alone reproduce both phase 1b Tmax medians
  (section above).
- **ELF compartment depletes plasma.** Not stated; chosen from the NONMEM
  general-linear meaning of `K14` / `K40` and because only this reading
  reproduces Tenero 2013's IV clearance (section above). With the
  non-depleting link reading the model would under-predict IV clearance by
  about 25%.
- **ELF concentration = ELF amount / Vc.** No ELF volume is printed; Vc is
  the only scaling that needs no extra parameter.
- **Logit-normal IIV on F.** The scale is not stated; logit is chosen
  because it reproduces the observed phase 1b AUC variability and keeps F
  below 1 (section above).
- **Allometric exponents.** Not printed. The parameter units (per 70 kg) and
  the "allometric scaling approach" in the Results point to the standard
  fixed exponents, 0.75 for CL, CLd and CLd2 and 1 for Vc, Vp and Vp2. K14,
  K40 and ka are reported in 1/h without a weight normalisation and are not
  scaled.
- **IIV on ka.** Table 2 prints the 29.5% IIV on the `Ka, fast` row only;
  one eta is shared by the fed and fasted ka, the usual NONMEM coding for a
  covariate-switched parameter.
- **Occasions.** The source does not define the two occasions of the IOV on
  ka. The simulations use occasion 1 for the first dosing day and 2 for all
  later days.
- **Food.** Food changes only the absorption rate in the model, so the
  phase 1b fed arm's lower AUC (26% below fasted) is not reproduced.
- **Virtual cohorts.** No demographics are reported. Phase 1b subjects are
  drawn with weight N(75, 12) kg truncated to 45-100 kg (the study admitted
  40-100 kg); Tenero 2013 subjects with N(80, 10) kg truncated to 55-110 kg
  (BMI 24-27 kg/m^2, 93% male).
- **No erratum** was found for the abstract or poster as of 2026-10-10.
