---
title: "Cyclosporine (Gao 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cyclosporine (Gao 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Gao_2022_cyclosporine"))$reference`
* Description: `r rxode2::rxode(readModelDb("Gao_2022_cyclosporine"))$description`
* Article: <https://doi.org/10.3389/fphar.2022.933739> (open access)

# Population

Gao 2022 pooled 681 sparse whole-blood cyclosporine concentrations from 157
Chinese children (age 1.5-18.8 years, median 7.8; body weight 12.0-91.0 kg,
median 27.5) with acquired aplastic anemia treated at two Shanghai hospitals
between 2014 and 2021 (Table 1). 59.9% had non-severe, 25.4% severe and
13.4% very severe disease; children who had received a stem-cell transplant
were excluded. Oral cyclosporine was started at 5 mg/kg/day in two divided
doses and titrated to 5-8 (up to 10) mg/kg/day against a 100-200 ng/mL trough
and 300-400 ng/mL peak window. Samples were therapeutic-drug-monitoring peaks
(2-4 h post-dose) and pre-dose troughs, measured with the Emit 2000
cyclosporine-specific assay. Median baseline total bilirubin was 10.90 umol/L
(range 4.00-70.00). Common co-medications were testosterone undecanoate
(70.7%), methylprednisolone (29.9%) and rabbit ATG (29.9%); none was retained
as a covariate. The sex distribution is not reported.

The same information is available programmatically via
`readModelDb("Gao_2022_cyclosporine")()$population`.

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` | log(1.26) 1/h | Table 2, Ka |
| `lcl` | log(29.1) L/h | Table 2, CL/F (typical 28-kg child) |
| `lvc` | log(325) L | Table 2, VC/F |
| `lq` | fixed(log(3.1)) L/h | Table 2, Q/F "FIX" |
| `lvp` | fixed(log(262)) L | Table 2, VP/F "FIX" |
| `e_wt_cl`, `e_wt_vc` | fixed 0.75, 1 | Eqs 2-3; Table 2 footnote |
| `e_tbili_cl` | -0.0107 per umol/L | Table 2, "TBIL on CL (%)" -1.07; footnote equation |
| `etalcl` | 0.0755 (28.0% CV) | Table 2, CV for IIV; log(CV^2 + 1) |
| `etalvc` | 0.326 (62.1% CV) | Table 2, CV for IIV; log(CV^2 + 1) |
| `expSd` | 0.348 | Table 2, sigma (additive on log concentrations) |
| Allometric reference 28 kg | n/a | Methods, Eqs 2-3 ("BWmedian ... (28 kg)") |
| TBIL centring 10.65 umol/L | n/a | Table 2 footnote equation |
| CL equation | n/a | Eq 2 and Table 2 footnote, with `1 +` restored (see below) |
| Two-compartment, first-order absorption ODEs | n/a | Results paragraph 2; Abstract |
| Log-additive residual error | n/a | Methods, Eq 1 paragraph |

# Virtual cohorts

Figure 3 of Gao 2022 simulates 1,000 children per body-weight band (10-60 kg)
at three fixed total-bilirubin levels. Here each band has 200 children. Panel A
is labelled "< 17.1 umol/L" (the upper limit of normal) without a single value;
it is simulated at the model's centring value, 10.65 umol/L. Panels B and C use
the printed 25.65 and 34.2 umol/L.

```{r cohort-fig3}
n_per_cell <- 200L
fig3_grid <- tidyr::expand_grid(
  panel = c("A", "B", "C"),
  WT = c(10, 20, 30, 40, 50, 60)
) |>
  dplyr::mutate(
    TBILI = c(A = 10.65, B = 25.65, C = 34.2)[panel],
    cell = dplyr::row_number()
  )

# One steady-state dose of 2.5 mg/kg (5 mg/kg/day split twice daily) and a
# trough observation one dosing interval later.
fig3_events <- fig3_grid |>
  dplyr::slice(rep(seq_len(dplyr::n()), each = n_per_cell)) |>
  dplyr::mutate(id = dplyr::row_number()) |>
  dplyr::reframe(
    dplyr::bind_rows(
      tibble::tibble(
        time = 0, evid = 1L, amt = 2.5 * WT, ss = 1L, ii = 12, cmt = "depot"
      ),
      tibble::tibble(
        time = 12, evid = 0L, amt = 0, ss = 0L, ii = 0, cmt = "central"
      )
    ),
    .by = c(id, panel, WT, TBILI, cell)
  ) |>
  dplyr::arrange(id, time, dplyr::desc(evid))
stopifnot(!anyDuplicated(fig3_events[, c("id", "time", "evid")]))
```

# Simulation

```{r simulate-fig3}
mod <- readModelDb("Gao_2022_cyclosporine")
rxode2::rxSetSeed(20220726)
sim_fig3 <- rxode2::rxSolve(
  mod,
  events = fig3_events,
  keep = c("panel", "WT", "TBILI"),
  maxsteps = 1e6
) |>
  as.data.frame()
stopifnot(!anyNA(sim_fig3$Cc))
```

# Replicate published figures

## Figure 3: steady-state troughs by body weight and total bilirubin

The published boxes (median and 25th-75th percentiles) were digitised by the
maintainers from the native-resolution Figure 3 image, with the y axis
calibrated on the 0-400 ng/mL gridlines (about 0.75 ng/mL per pixel). The
boxes show exposure without residual error, so the comparison uses `Cc`.

```{r fig3-published}
fig3_published <- tibble::tribble(
  ~panel, ~WT, ~q25, ~q50, ~q75,
  "A", 10, 48.5, 79.9, 116.4,
  "A", 20, 64.9, 99.3, 142.2,
  "A", 30, 78.0, 117.9, 159.0,
  "A", 40, 88.4, 132.1, 180.6,
  "A", 50, 101.9, 145.5, 196.3,
  "A", 60, 103.7, 154.5, 207.5,
  "B", 10, 70.5, 103.0, 146.3,
  "B", 20, 89.6, 136.6, 187.3,
  "B", 30, 106.7, 157.5, 207.5,
  "B", 40, 126.9, 177.6, 234.3,
  "B", 50, 133.2, 186.6, 239.9,
  "B", 60, 144.8, 201.5, 259.0,
  "C", 10, 84.0, 125.4, 171.6,
  "C", 20, 116.4, 161.2, 217.5,
  "C", 30, 131.0, 176.9, 241.0,
  "C", 40, 147.0, 197.8, 269.4,
  "C", 50, 161.2, 218.7, 282.5,
  "C", 60, 164.9, 231.0, 303.0
)

fig3_sim <- sim_fig3 |>
  dplyr::filter(time == 12) |>
  dplyr::summarise(
    q25 = quantile(Cc, 0.25),
    q50 = quantile(Cc, 0.50),
    q75 = quantile(Cc, 0.75),
    .by = c(panel, WT)
  )

fig3_cmp <- dplyr::inner_join(
  fig3_sim,
  fig3_published,
  by = c("panel", "WT"),
  suffix = c("_sim", "_pub")
) |>
  dplyr::mutate(
    pct_diff_median = 100 * (q50_sim / q50_pub - 1),
    pct_diff_iqr = 100 * ((q75_sim - q25_sim) / (q75_pub - q25_pub) - 1)
  )
stopifnot(nrow(fig3_cmp) == 18L)

fig3_cmp |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), \(x) round(x, 1))) |>
  dplyr::select(
    panel, WT, q50_sim, q50_pub, pct_diff_median, pct_diff_iqr
  ) |>
  dplyr::rename(
    "Panel" = panel,
    "Weight (kg)" = WT,
    "Simulated median (ng/mL)" = q50_sim,
    "Published median (ng/mL)" = q50_pub,
    "Median difference (%)" = pct_diff_median,
    "IQR-width difference (%)" = pct_diff_iqr
  ) |>
  knitr::kable(caption = "Steady-state trough, simulated vs. Figure 3.")
```

```{r fig3-plot, fig.width = 8, fig.height = 4}
panel_labels <- c(
  A = "(A) TBIL < 17.1 umol/L (sim. 10.65)",
  B = "(B) TBIL = 25.65 umol/L",
  C = "(C) TBIL = 34.2 umol/L"
)
ggplot() +
  geom_boxplot(
    data = sim_fig3 |> dplyr::filter(time == 12),
    aes(x = factor(WT), y = Cc),
    outlier.shape = NA, coef = 0, fill = "grey90"
  ) +
  geom_point(
    data = fig3_published,
    aes(x = factor(WT), y = q50),
    colour = "firebrick", shape = 4, size = 3
  ) +
  geom_errorbar(
    data = fig3_published,
    aes(x = factor(WT), ymin = q25, ymax = q75),
    colour = "firebrick", width = 0.3
  ) +
  facet_wrap(~panel, labeller = labeller(panel = panel_labels)) +
  coord_cartesian(ylim = c(0, 400)) +
  labs(
    x = "Body weight (kg)",
    y = "Cyclosporine trough (ng/mL)",
    caption = paste(
      "Replicates Figure 3 of Gao 2022. Grey boxes: simulated median and IQR;",
      "red: digitised published median and IQR."
    )
  )
```

The dose (mg/kg), the clearance, both volumes and the bilirubin slope all set
the median trough, and the two IIV variances set the box height. A
mis-transcribed value in any of them moves these columns by tens of percent.
Each simulated median has a Monte-Carlo SE of about 5% (log-scale SD about 0.55
over 200 children), so the gate is on the centre and the 90th percentile of
the 18 cells, never on the worst cell.

```{r fig3-gate}
# Over five seeds: median -0.8 to 1.5%, 90th percentile 5.7-6.9%, IQR-width
# median -5.5 to 1.0%. A 20% error in CL/F moves the median trough ~25%.
stopifnot(
  abs(median(fig3_cmp$pct_diff_median)) < 6,
  quantile(abs(fig3_cmp$pct_diff_median), 0.9) < 15,
  abs(median(fig3_cmp$pct_diff_iqr)) < 15
)
```

## Figure 2: visual predictive check

Gao 2022 Figure 2 is a prediction-corrected VPC of the observed data, whose
doses were titrated between 5 and 10 mg/kg/day. The simulation below is a
qualitative counterpart. It uses 200 children at the starting regimen of
2.5 mg/kg twice daily, with weight and bilirubin drawn log-normally around the
Table 1 medians, and includes residual error (`sim`). The red points are the
observed 5th, 50th and 95th percentiles read by eye from Figure 2, so they are
approximate.

```{r cohort-vpc}
set.seed(933739)
n_vpc <- 200L
vpc_subjects <- tibble::tibble(
  id = seq_len(n_vpc),
  WT = pmin(pmax(exp(rnorm(n_vpc, log(27.5), 0.40)), 12), 91),
  TBILI = pmin(pmax(exp(rnorm(n_vpc, log(10.9), 0.45)), 4), 70)
)
vpc_times <- c(seq(0, 4, by = 0.25), seq(4.5, 12, by = 0.5))
vpc_events <- vpc_subjects |>
  dplyr::reframe(
    dplyr::bind_rows(
      tibble::tibble(
        time = 0, evid = 1L, amt = 2.5 * WT, ss = 1L, ii = 12, cmt = "depot"
      ),
      tibble::tibble(
        time = vpc_times, evid = 0L, amt = 0, ss = 0L, ii = 0, cmt = "central"
      )
    ),
    .by = c(id, WT, TBILI)
  ) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim_vpc <- rxode2::rxSolve(
  mod,
  events = vpc_events,
  keep = c("WT", "TBILI"),
  maxsteps = 1e6
) |>
  as.data.frame()
stopifnot(!anyNA(sim_vpc$Cc))
```

```{r fig2-plot, fig.width = 6, fig.height = 4}
fig2_observed <- tibble::tribble(
  ~time, ~p05, ~p50, ~p95,
  2, 57, 230, 1450,
  3, 120, 380, 900,
  4, 115, 255, 620,
  11, 39, 110, 270,
  11.5, 39, 70, 195,
  12, 33, 80, 180
) |>
  tidyr::pivot_longer(-time, names_to = "percentile", values_to = "conc")

sim_vpc |>
  dplyr::summarise(
    p05 = quantile(sim, 0.05),
    p50 = quantile(sim, 0.50),
    p95 = quantile(sim, 0.95),
    .by = time
  ) |>
  ggplot(aes(time)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.25) +
  geom_line(aes(y = p50)) +
  geom_point(
    data = fig2_observed,
    aes(y = conc, shape = percentile),
    colour = "firebrick"
  ) +
  scale_y_log10(limits = c(10, 2000)) +
  labs(
    x = "Time after dose (h)",
    y = "Cyclosporine (ng/mL)",
    shape = "Observed",
    caption = paste(
      "Compare with Figure 2 of Gao 2022. Band: simulated 5th-95th",
      "percentiles at 5 mg/kg/day; line: median."
    )
  )
```

# PKNCA validation

Gao 2022 reports no NCA table. PKNCA is run over one steady-state dosing
interval of the VPC cohort. The check is cohort-free: each child's AUCtau at
steady state must equal Dose / CL for that child's own clearance (`cl` in the
simulation output).

```{r pknca}
sim_nca <- sim_vpc |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(treatment = "2.5 mg/kg q12h") |>
  dplyr::select(id, time, Cc, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- vpc_events |>
  dplyr::filter(evid == 1) |>
  dplyr::mutate(treatment = "2.5 mg/kg q12h") |>
  dplyr::select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0,
  end = 12,
  cmax = TRUE,
  tmax = TRUE,
  cmin = TRUE,
  auclast = TRUE
)
nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

auc_check <- sim_vpc |>
  dplyr::distinct(id, cl, WT) |>
  dplyr::inner_join(nca_wide, by = "id") |>
  dplyr::mutate(
    auc_expected = 1000 * 2.5 * WT / cl,
    pct_diff = 100 * (auclast / auc_expected - 1)
  )
stopifnot(nrow(auc_check) == n_vpc)

auc_check |>
  dplyr::summarise(
    dplyr::across(c(cmax, tmax, cmin, auclast), median)
  ) |>
  dplyr::mutate(dplyr::across(dplyr::everything(), \(x) signif(x, 3))) |>
  dplyr::rename(
    "Cmax,ss (ng/mL)" = cmax,
    "Tmax (h)" = tmax,
    "Cmin,ss (ng/mL)" = cmin,
    "AUCtau,ss (ng*h/mL)" = auclast
  ) |>
  knitr::kable(caption = "Median steady-state NCA, 2.5 mg/kg twice daily.")
```

AUCtau and Dose/CL use the same drawn parameters, so the only difference is
trapezoidal error on the 0.25-0.5 h grid around the peak. The peak is
broad (Ka 1.26 1/h), so that error is a fraction of a percent.

```{r pknca-gate}
# Realised maximum 0.27-0.42% over five seeds.
stopifnot(max(abs(auc_check$pct_diff)) < 2)
```

The paper's Results note that children under 30 kg fall below the
100-200 ng/mL trough window on 5 mg/kg/day. In the Figure 3 replication, the
simulated median trough at 10 and 20 kg with normal bilirubin is below
100 ng/mL, which agrees.

# Assumptions and deviations

- **Bilirubin equation.** The Table 2 footnote prints
  `CL = CLtypical x ((TBIL - 10.65) x -0.0107)`, which would make clearance
  negative at the reference bilirubin. The maintainers restored the missing
  `1 +`: `CL = CLtypical x (1 - 0.0107 x (TBIL - 10.65))`. This matches the
  text ("a 1.07% reduction per 1 ... rise in total bilirubin"), and the
  Figure 3 replication above reproduces all three bilirubin panels with it.
  The Abstract and Discussion write "per 1 nmol/L"; Table 1 reports bilirubin
  in umol/L and the Figure 3 thresholds (17.1, 25.65, 34.2 umol/L) are in
  umol/L, so the slope is per umol/L. The linear factor reaches zero at
  TBIL = 104 umol/L, above the cohort maximum of 70 umol/L. The model does not
  truncate it.
- **Fixed Q/F and Vp/F.** The Results text says Vp/F and Q/F were fixed at
  "496 and 5 L/h" from Eljebari et al. 2012, while Table 2 prints 262 L and
  3.1 L/h "FIX". Table 2 gives estimates for the typical 28-kg child, and both
  Table 2 values equal the text values times the same allometric weight ratio
  (262/496 = 0.528 and 3.1/5 = 0.620 = 0.528^0.75). So the text quotes the
  literature values at the literature model's reference weight, and Table 2
  quotes them scaled to 28 kg. The Discussion's total V/F normalised to 70 kg,
  1,468 L, equals (325 + 262) x 70/28 = 1,467.5 L, which confirms 262 L. The
  model uses the Table 2 values.
- **IIV scale.** The Table 2 footnote gives the IIV CV as
  "100 x (e^variance)^1/2". Read literally this is always above 100%, so the
  maintainers read it as the log-normal CV, 100 x sqrt(e^variance - 1), and
  used omega^2 = log(CV^2 + 1). The other reading (CV = 100 x omega) gives
  omega^2 = 0.0784 and 0.386 for CL/F and Vc/F. Figure 3 cannot tell the two
  apart: at 1,000 children per band both reproduce the published IQRs to
  within a few percent. The SIR intervals are percentile intervals, which the
  scale change does not affect.
- **Residual error.** Table 2 labels the value 0.348 "sigma", and the Methods
  define the residual variance as sigma^2, so 0.348 is used as the log-scale
  SD (`lnorm(expSd)`). The printed SIR interval (0.356-0.393) does not contain
  the point estimate and was not used.
- **Allometric reference.** The Methods give the median weight as 28 kg and
  use it for centring, although Table 1 prints 27.5 kg. The model uses 28 kg.
- **Figure 3 panel A** has no single bilirubin value ("< 17.1 umol/L") and is
  simulated at 10.65 umol/L, the centring value.
- **Figure 2 cohort.** Individual doses, weights and bilirubin values are not
  available. The VPC cohort draws weight and bilirubin log-normally around the
  Table 1 medians, clipped to the Table 1 ranges, and doses everyone at the
  starting regimen. The observed percentiles were read by eye, so Figure 2 is
  a qualitative comparison only.
- No erratum or correction notice for Gao 2022 was found in Europe PMC or
  Crossref as of 2026-10-05.
