---
title: "Cefquinome PK/PD against Escherichia coli in foals (Gao 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cefquinome PK/PD against Escherichia coli in foals (Gao 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Gao 2025 is a veterinary pharmacokinetic/pharmacodynamic integration study of
**cefquinome**, a fourth-generation cephalosporin licensed only for animal use,
against *Escherichia coli* isolated from a septicaemic foal. Ten Ili foals
received 1 mg/kg intravenously and intramuscularly in a two-phase crossover;
serum drawn after the intramuscular dose was then used *ex vivo* to build
time-kill curves, and the resulting effect was integrated against the PK/PD
index AUC0-24h/MIC divided by 24 h.

The paper has two separable quantitative components, and this package carries
one model file for each.

* Article: <https://doi.org/10.3390/vetsci12040294>

```{r models}
mods <- c("Gao_2025_cefquinome_foal", "Gao_2025_cefquinome_pkpd_index")
uis <- lapply(mods, function(m) rxode2::rxode(readModelDb(m)))
names(uis) <- mods
cat(uis[[1]]$reference, "\n")
```

| Model file | Paper component | Source |
|---|---|---|
| `Gao_2025_cefquinome_foal` | One-compartment PK with first-order intramuscular absorption | Table 1, Figure 2 |
| `Gao_2025_cefquinome_pkpd_index` | Inhibitory sigmoid Emax PK/PD integration | Section 2.9, Table 4, Figure 6 |

```{r descriptions, results = "asis"}
for (m in mods) cat("**", m, "** -- ", uis[[m]]$description, "\n\n", sep = "")
```

# Population

Ten healthy Ili foals aged 7 months to 1 year, mean weight 191 +/- 21.7 kg,
supplied by Zhaosu County Xiyu Horse Industry Co., Ltd. (Yili, Xinjiang,
China). All animals were acclimatised for two weeks, passed a fitness
assessment (physical examination, heart rate, rectal temperature) and had
received no medical treatment for at least two weeks before enrolment.

The design was a two-phase crossover: five foals received 1 mg/kg cefquinome
intramuscularly in the cervical region and five received 1 mg/kg intravenously
via the jugular vein; after a two-week washout the routes were reversed. Blood
was drawn from the left jugular vein at 0.083, 0.167, 0.25, 0.5, 0.75, 1, 2, 3,
6, 9, 12 and 24 h and serum assayed by HPLC (LLOQ 0.01 ug/mL). Baseline
characteristics are given in Gao 2025 Section 2.2; the PK results are Table 1.

The pharmacodynamic component used *E. coli* strain HE13, isolated from a foal
with septicaemia in Heilongjiang Province, plus a 53-isolate MIC survey
(Figure 3) and a 20-isolate broth-versus-serum comparison (Table 2).

The same information is available programmatically:

```{r population}
str(uis[["Gao_2025_cefquinome_foal"]]$population, max.level = 1)
```

# Source trace

Every `ini()` entry carries an in-file comment naming its source location. The
table below collects them, and flags the two values that Gao 2025 does **not**
print and that are therefore derived here from published Table 1 statistics by
standard one-compartment identities.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` (CL) | 0.09 L/h/kg | Table 1, IV column |
| `lvc` (vc) | 0.3051 L/kg | **Derived**: CL / (log(2) / T1/2beta) from Table 1, IV column |
| `lka` (ka) | 0.6851 1/h | **Derived**: root of log(ka/kel)/(ka - kel) = Tmax, Table 1 IM Tmax = 2.16 h |
| `lfdepot` (F) | 0.4386 | Table 1, IM column, F = 43.86% |
| `addSd` | 0 (fixed) | Not reported; held at zero for typical-value simulation |
| PK structure `d/dt(depot)`, `d/dt(central)` | n/a | Not published; one-compartment structure imposed, see Assumptions |
| `e0` | 2.88 log10 CFU/mL | Table 4, printed as "Emax" (Section 2.9 defines it as the no-drug change) |
| `emax` | -4.65 log10 CFU/mL | Table 4, printed as "E0" (Section 2.9 defines it as the drug-treated change) |
| `lec50` (EC50) | 2.61 | Table 4 |
| `lhill` (N) | 4.22 | Table 4 |
| `mic` | 0.062 ug/mL | Section 3.3, strain HE13 in foal serum; also the 53-isolate MIC50 |
| `log10_cfu0` | 6 | Section 2.8, ex vivo inoculum ~1 x 10^6 CFU/mL |
| Sigmoid equation | n/a | Section 2.9 |
| Dose equation | n/a | Section 2.11 |

# Part 1 - Pharmacokinetics of cefquinome in foals

Gao 2025 analysed serum concentrations non-compartmentally in WinNonlin 5.2.1
and published no compartmental model. The packaged model is the smallest
structure that the published Table 1 statistics determine: one compartment,
first-order intramuscular absorption, and a bioavailability applied to the
depot input only.

```{r pk-events}
pk_mod <- readModelDb("Gao_2025_cefquinome_foal")

grid <- seq(0, 24, by = 0.02)

pk_events <- dplyr::bind_rows(
  # Intravenous arm: dose the `central` compartment directly.
  data.frame(id = 1L, time = 0, amt = 1, evid = 1L, cmt = "central"),
  data.frame(id = 1L, time = grid, amt = NA_real_, evid = 0L, cmt = "central"),
  # Intramuscular arm: dose the `depot` compartment so f(depot) applies.
  data.frame(id = 2L, time = 0, amt = 1, evid = 1L, cmt = "depot"),
  data.frame(id = 2L, time = grid, amt = NA_real_, evid = 0L, cmt = "central")
) |>
  dplyr::mutate(arm = ifelse(id == 1L, "IV", "IM")) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

pk_sim <- rxode2::rxSolve(pk_mod, events = pk_events, keep = "arm") |>
  as.data.frame()

stopifnot(length(unique(pk_sim$id)) == 2L, all(is.finite(pk_sim$Cc)), all(pk_sim$Cc >= 0))
```

## Replicating Figure 2

```{r figure-2}
# Replicates Figure 2 of Gao 2025: semi-logarithmic serum concentration versus
# time after a single 1 mg/kg IV or IM injection. Gao 2025 plots to 15 h; the
# last quantifiable observation in that figure is at 12 h, at about
# 0.09 ug/mL for both routes.
pk_sim |>
  dplyr::filter(Cc > 0.01) |>
  ggplot(aes(time, Cc, colour = arm)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0.01, linetype = "dotted") +
  scale_y_log10(limits = c(0.01, 10)) +
  coord_cartesian(xlim = c(0, 15)) +
  labs(x = "Time (h)", y = "Serum concentration (ug/mL)", colour = "Route",
       title = "Figure 2 - serum cefquinome after 1 mg/kg",
       caption = "Replicates Figure 2 of Gao 2025. Dotted line: 0.01 ug/mL LLOQ.")
```

The 12 h concentration is the discriminator that fixes the volume of
distribution. Gao 2025 reports both a terminal half-life (2.35 h) and a mean
residence time (2.67 h) for the intravenous route, and these imply *different*
one-compartment volumes. Figure 2 settles which is right.

```{r volume-discriminator}
cl_pub  <- 0.09    # Table 1, IV
t12_pub <- 2.35    # Table 1, IV
mrt_pub <- 2.67    # Table 1, IV
kel     <- log(2) / t12_pub

vol_cand <- tibble::tibble(
  Derivation = c("vc = CL / kel (packaged)", "vc = CL x MRT0-last (rejected)"),
  vc         = c(cl_pub / kel, cl_pub * mrt_pub)
) |>
  dplyr::mutate(
    `IV C(12 h), ug/mL` = (1 / vc) * exp(-(cl_pub / vc) * 12),
    `Figure 2 reads`    = "~0.09"
  )

knitr::kable(vol_cand, digits = 4,
             caption = "Only vc = CL / kel reproduces the 12 h intravenous concentration in Figure 2.")
```

The two derivations disagree because the intravenous profile in Figure 2 is
visibly biexponential over the first two to three hours, so `MRT0-last` (2.67 h)
is shorter than `1/kel` (3.39 h). A two-compartment model needs four
disposition parameters and Gao 2025 publishes only three independent
intravenous summaries, so the distribution phase is not identifiable and is not
modelled.

## The intramuscular absorption rate constant, and why it is not flip-flop

`ka` is derived from the reported `Tmax`. The alternative that would instead
reproduce the reported intramuscular `T1/2beta` of 4.16 h is a flip-flop
parameterisation, and it is refuted by two other published numbers.

```{r ka-falsifier}
tmax_pub <- 2.16
cmax_pub <- 0.89
fbio     <- 0.4386

# Use the PACKAGED volume, so every check below tests the shipped model rather
# than a re-derivation of it. The packaged value is the derivation rounded to
# four decimal places; pin that rounding rather than let it drift.
pk_theta   <- uis[["Gao_2025_cefquinome_foal"]]$theta
vc         <- exp(pk_theta[["lvc"]])
vc_derived <- cl_pub / kel
stopifnot(abs(vc - vc_derived) < 5e-5)

tmax_of <- function(ka) log(ka / kel) / (ka - kel)
cmax_of <- function(ka) (fbio / vc) * exp(-kel * tmax_of(ka))

ka_fit <- uniroot(function(ka) tmax_of(ka) - tmax_pub, c(kel * 1.001, 20))$root
ka_ff  <- log(2) / 4.16

ka_cmp <- tibble::tibble(
  Parameterisation = c("ka from Tmax (packaged)", "flip-flop, ka from IM T1/2beta"),
  `ka (1/h)`       = c(ka_fit, ka_ff),
  `Tmax (h)`       = c(tmax_of(ka_fit), tmax_of(ka_ff)),
  `Cmax (ug/mL)`   = c(cmax_of(ka_fit), cmax_of(ka_ff))
) |>
  dplyr::mutate(
    `Published Tmax (h)`     = tmax_pub,
    `Published Cmax (ug/mL)` = cmax_pub
  )

knitr::kable(ka_cmp, digits = 4,
             caption = "The flip-flop root doubles Tmax and underpredicts Cmax 2.3-fold.")

# The packaged value must match the derivation, and Cmax must land inside the
# published SD of 0.14 without ever having been fitted to it.
stopifnot(
  abs(exp(pk_theta[["lka"]]) - ka_fit) < 5e-4,
  abs(cmax_of(ka_fit) - cmax_pub) < 0.14
)
```

The packaged `ka` reproduces `Tmax` by construction, but `Cmax` is an
*independent* check: it was not used in the derivation, and the identity
`Cmax = (F x Dose / vc) x exp(-kel x Tmax)` returns
`r sprintf("%.3f", cmax_of(ka_fit))` ug/mL against the published 0.89 +/- 0.14
ug/mL. The flip-flop root returns `r sprintf("%.3f", cmax_of(ka_ff))` ug/mL.

## PKNCA validation against Table 1

Gao 2025 reports `AUC0-last` and `MRT0-last`. Figure 2 shows the last
quantifiable sample at 12 h -- with a 2.35 h terminal half-life the 12-24 h
contribution falls below the 0.01 ug/mL limit of quantification -- so the NCA
window below ends at 12 h to score the same quantity the paper reports.

```{r pknca}
sim_nca <- pk_sim |>
  dplyr::filter(!is.na(Cc), time <= 12) |>
  dplyr::select(id, time, Cc, arm)

# PKNCA needs a time-zero record per subject to anchor AUC0-*. The simulation
# grid already starts at 0, so no defensive row is added -- and none should be:
# for the IV bolus arm the time-zero value is the post-dose C0 = Dose/vc, which
# an unconditional "Cc = 0 at time 0" row would silently destroy. Assert the
# rows are there and carry the right values instead of assuming it.
t0 <- sim_nca |> dplyr::filter(time == 0)
stopifnot(
  nrow(t0) == 2L,
  # IV bolus: the time-zero record is the post-dose C0 = Dose / vc.
  abs(t0$Cc[t0$arm == "IV"] - 1 / vc) < 1e-6,
  # Extravascular: pre-dose concentration is exactly zero.
  t0$Cc[t0$arm == "IM"] == 0
)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

dose_df <- pk_events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, arm)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)

intervals <- data.frame(
  start = 0, end = 12,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r nca-compare}
# Gao 2025 Table 1 (mean +/- SD over 10 foals). Cmax and Tmax are printed as
# "--" for the intravenous arm and are left NA here rather than invented.
published <- tibble::tribble(
  ~arm, ~cmax, ~tmax, ~auclast, ~half.life,
  "IV",   NA,    NA,    12.33,    2.35,
  "IM",   0.89,  2.16,   5.41,    4.16
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by        = "arm",
  units     = c(cmax = "ug/mL", tmax = "h", auclast = "h*ug/mL", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(cmp, caption = "Simulated versus Gao 2025 Table 1. * differs by >20%.")
```

Both `AUC0-last` values run about 10-15% below the published means. That gap is
structural, not a transcription error: `CL` and `AUC0-last` in Table 1 are
separately averaged per-foal non-compartmental estimates, so `Dose / CL`
(11.1 h\*ug/mL intravenously) does not equal the mean `AUC0-last`
(12.33 h\*ug/mL). Because the model takes `CL` as its primary, it inherits the
`Dose / CL` side of that discrepancy. This is the usual mean-of-ratios versus
ratio-of-means gap.

The one starred row is the intramuscular half-life: PKNCA's terminal fit over
the 0-12 h window returns 2.44 h against the published 4.16 h. (The model's
disposition half-life is 2.35 h exactly, as the intravenous row confirms; the
extra 0.09 h is residual absorption still contaminating the terminal slope at
12 h, not a second disposition phase.) The gap to 4.16 h is the deliberate
consequence of rejecting the flip-flop parameterisation above, and it is
supported by Figure 2, in which the two routes have visibly parallel terminal
slopes rather than the shallower intramuscular slope flip-flop absorption would
produce.

# Part 2 - Ex vivo PK/PD integration

Gao 2025 Section 2.9 fits

E = Emax - (Emax - E0) x Ce^N / (Ce^N + EC50^N)

with the paper's `Emax` defined as the 24 h change in the **drug-free** samples
(+2.88) and its `E0` as the change in the **drug-treated** samples (-4.65) --
the reverse of the usual convention. The packaged model carries the same two
numbers under canonical names (`e0` = 2.88, `emax` = -4.65) and the
algebraically identical form.

## Back-solving the published targets

Gao 2025 Table 4 also prints the index values required for bacteriostatic,
bactericidal and bacterial-elimination action. Those are printed independently
of the four sigmoid parameters, so recovering them from the fitted curve is a
genuine check on the transcription and the equation form.

```{r pd-targets}
pd_ui <- uis[["Gao_2025_cefquinome_pkpd_index"]]
th    <- pd_ui$theta

e0    <- th[["e0"]]
emax  <- th[["emax"]]
ec50  <- exp(th[["lec50"]])
hill  <- exp(th[["lhill"]])
mic   <- th[["mic"]]

effect_of <- function(ce) e0 - (e0 - emax) * ce^hill / (ce^hill + ec50^hill)
index_for <- function(target_e) {
  uniroot(function(ce) effect_of(ce) - target_e, c(1e-6, 1e4))$root
}

targets <- tibble::tibble(
  Endpoint = c("Bacteriostatic", "Bactericidal", "Bacterial elimination"),
  `E (log10 CFU/mL)` = c(0, -3, -4),
  `Index, back-solved` = vapply(c(0, -3, -4), index_for, numeric(1)),
  `Index, Table 4` = c(2.34, 3.53, 4.86)
) |>
  dplyr::mutate(`% diff` = 100 * (`Index, back-solved` - `Index, Table 4`) / `Index, Table 4`)

knitr::kable(targets, digits = 3,
             caption = "Back-solved versus published PK/PD index targets (Gao 2025 Table 4).")

# The bacteriostatic and bactericidal targets must reproduce tightly; they are
# the evidence that e0/emax/EC50/N were transcribed correctly.
stopifnot(
  abs(index_for(0)  - 2.34) < 0.02,
  abs(index_for(-3) - 3.53) < 0.02
)
```

The first two targets reproduce to within 0.5%. The third does not: the fitted
curve places a 4 log10 reduction at
`r sprintf("%.3f", index_for(-4))`, not at the published 4.86. Running it the
other way, an index of 4.86 corresponds to
`r sprintf("%.3f", effect_of(4.86))` log10 CFU/mL. Because the other two targets
land almost exactly on the round definitions of 0 and -3, the published 4.86
appears to be internally inconsistent with Table 4's own sigmoid parameters. See
the Errata.

## Replicating Figure 6

```{r figure-6}
# Replicates Figure 6 of Gao 2025: inhibitory sigmoid Emax model relating the
# ex vivo AUC0-24h/MIC divided by 24 h to the 24 h change in bacterial count.
tibble::tibble(ce = seq(0, 8, length.out = 400)) |>
  dplyr::mutate(E = effect_of(ce)) |>
  ggplot(aes(ce, E)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = c(0, -3), linetype = "dashed", colour = "grey50") +
  geom_point(
    data = targets |> dplyr::rename(ce = `Index, Table 4`, E = `E (log10 CFU/mL)`),
    aes(ce, E), colour = "firebrick", size = 2.5
  ) +
  labs(x = "AUC0-24h / MIC divided by 24 h", y = "Change in count over 24 h (log10 CFU/mL)",
       title = "Figure 6 - inhibitory sigmoid Emax model",
       caption = paste("Replicates Figure 6 of Gao 2025. Red points are the Table 4 targets;",
                       "the elimination point sits off the curve (see Errata)."))
```

## The packaged ODE reproduces the per-interval change

The model spreads each interval's change uniformly across 24 h, so
`log10(bact)` must move by exactly `E` over one interval.

```{r pd-ode-check}
pd_mod <- readModelDb("Gao_2025_cefquinome_pkpd_index")

# One "subject" per index level; AUC_CEFQ is chosen so that
# AUC_CEFQ / mic / 24 equals the wanted index.
levels_ce <- c(0, 1, 2.34, 3.53, 4.86, 8)
pd_events <- do.call(rbind, lapply(seq_along(levels_ce), function(i) {
  data.frame(
    id       = i,
    time     = c(0, 24),
    amt      = NA_real_,
    evid     = 0L,
    cmt      = "bact",
    AUC_CEFQ = levels_ce[i] * mic * 24,
    ce_level = levels_ce[i]
  )
}))

pd_sim <- rxode2::rxSolve(pd_mod, events = pd_events, keep = c("AUC_CEFQ", "ce_level")) |>
  as.data.frame()

ode_chk <- pd_sim |>
  dplyr::group_by(ce_level) |>
  dplyr::summarise(
    # Measure the STATE, not the observable: Cc is log10(bact + 1), and that
    # 1-CFU/mL floor deliberately shifts the reading once bact is driven into
    # the tens of CFU/mL. The floor is a guard on the observation, not part of
    # the kinetics, so the ODE must be scored on bact itself.
    `Delta log10 (state)` = log10(bact[time == 24]) - log10(bact[time == 0]),
    `Delta log10 (Cc, floored)` = Cc[time == 24] - Cc[time == 0],
    .groups = "drop"
  ) |>
  dplyr::mutate(`Sigmoid E` = effect_of(ce_level),
                `abs diff`  = abs(`Delta log10 (state)` - `Sigmoid E`))

knitr::kable(ode_chk, digits = 5,
             caption = paste("The ODE moves log10(bact) by exactly the sigmoid's E over each 24 h",
                             "interval. The floored observable diverges only at maximal kill,",
                             "where bact falls to tens of CFU/mL."))

stopifnot(max(ode_chk$`abs diff`) < 1e-5)
```

## Ex vivo time-kill curves (Figure 5)

In an *ex vivo* time-kill the serum sample is held at a fixed concentration for
the 24 h incubation, so that sample's exposure is `C(t_draw) x 24 h` and the
index reduces to `C(t_draw) / MIC`. Feeding the Part 1 PK model into the Part 2
PD model therefore reproduces Figure 5 directly.

```{r figure-5}
draw_times <- c(0, 0.083, 0.167, 0.25, 0.5, 0.75, 1, 2, 3, 6, 9, 12, 24)

conc_at_draw <- pk_sim |>
  dplyr::filter(arm == "IM") |>
  dplyr::select(time, Cc)

c_of <- stats::approxfun(conc_at_draw$time, conc_at_draw$Cc)

exvivo_events <- do.call(rbind, lapply(seq_along(draw_times), function(i) {
  data.frame(
    id       = i,
    time     = seq(0, 24, by = 0.5),
    amt      = NA_real_,
    evid     = 0L,
    cmt      = "bact",
    AUC_CEFQ = c_of(draw_times[i]) * 24,
    draw     = draw_times[i]
  )
}))

exvivo <- rxode2::rxSolve(pd_mod, events = exvivo_events, keep = c("AUC_CEFQ", "draw")) |>
  as.data.frame()

# Replicates Figure 5 of Gao 2025: ex vivo antibacterial activity of cefquinome
# against E. coli in serum drawn at each sampling time after IM administration.
exvivo |>
  dplyr::mutate(draw = factor(draw)) |>
  ggplot(aes(time, Cc, colour = draw)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = log10(200), linetype = "dotted") +
  labs(x = "Incubation time (h)", y = "log10 CFU/mL", colour = "Serum draw (h)",
       title = "Figure 5 - ex vivo time-kill in foal serum",
       caption = "Replicates Figure 5 of Gao 2025. Dotted line: 200 CFU/mL detection limit.")
```

Gao 2025 Section 3.6 states that "a rapid bactericidal activity was noted in
serum samples collected between 0.75 h and 8 h". The packaged model puts the
bactericidal threshold (index 3.53, i.e. a serum concentration of
`r sprintf("%.3f", 3.53 * mic)` ug/mL) over the following window:

```{r bactericidal-window}
im_prof <- pk_sim |> dplyr::filter(arm == "IM")
thr     <- index_for(-3) * mic
above   <- im_prof$time[im_prof$Cc >= thr]

cat(sprintf("Serum stays at or above %.3f ug/mL from %.2f h to %.2f h\n",
            thr, min(above), max(above)))
cat("Gao 2025 Section 3.6 reports rapid bactericidal activity from 0.75 h to 8 h\n")
```

The upper edge of the window is an independent corroboration of the PK model:
the paper's stated 8 h and the model's
`r sprintf("%.1f", max(pk_sim$time[pk_sim$arm == "IM" & pk_sim$Cc >= index_for(-3) * mic]))`
h agree without either having been fitted to the other.

# Part 3 - Monte Carlo target attainment

Gao 2025 Section 2.10 ran a 10,000-subject Monte Carlo in Crystal Ball,
treating `AUC0-24h` as log-normal and sampling the 53-isolate MIC distribution
of Figure 3, and reported target attainment rates of **46.38%** (bactericidal)
and **24.48%** (bacterial elimination) at the recommended 1 mg/kg intramuscular
dose. That analysis is entirely separate from the sigmoid fit, so reproducing it
tests the whole chain at once.

```{r monte-carlo}
set.seed(20250322)

# Figure 3: MIC distribution of the 53 serum isolates.
mic_dist <- tibble::tibble(
  mic = c(0.016, 0.031, 0.062, 0.125, 0.25, 0.5, 1),
  n   = c(4L, 9L, 20L, 9L, 7L, 2L, 2L)
)
stopifnot(sum(mic_dist$n) == 53L)

# Table 1, IM column: AUC0-last 5.41 +/- 0.81 h*ug/mL, taken as log-normal.
auc_mean <- 5.41
auc_sd   <- 0.81
sdlog    <- sqrt(log(1 + (auc_sd / auc_mean)^2))
meanlog  <- log(auc_mean) - 0.5 * sdlog^2

n_sim <- 10000L
draws <- tibble::tibble(
  auc = rlnorm(n_sim, meanlog, sdlog),
  mic = sample(mic_dist$mic, n_sim, replace = TRUE, prob = mic_dist$n / sum(mic_dist$n))
) |>
  dplyr::mutate(index = auc / mic / 24)

tar <- tibble::tibble(
  Endpoint = c("Bactericidal", "Bacterial elimination"),
  `Index target (Table 4)` = c(3.53, 4.86),
  `TAR simulated (%)` = c(
    100 * mean(draws$index >= 3.53),
    100 * mean(draws$index >= 4.86)
  ),
  `TAR published (%)` = c(46.38, 24.48)
) |>
  dplyr::mutate(`Difference (pp)` = `TAR simulated (%)` - `TAR published (%)`)

knitr::kable(tar, digits = 2,
             caption = "Monte Carlo target attainment at 1 mg/kg IM, versus Gao 2025 Section 3.8.")

stopifnot(all(abs(tar$`Difference (pp)`) < 4))
```

Both rates reproduce to within a couple of percentage points of the published
values, from the Figure 3 MIC distribution and the Table 1 intramuscular AUC
alone.

# Part 4 - Dose regimen calculation

Gao 2025 Section 2.11 computes the dose required for each endpoint as

Dose = factor x MIC x Clearance(per day) / (fu x F)

where `factor` is the index target, `fu` is the unbound fraction, and `F` is
the absolute bioavailability. The paper reports doses of 1.10, 1.66 and
2.28 mg/kg, but **never states a value for `fu`**, so the equation cannot be
evaluated as printed. Because the three published doses are exactly
proportional to the three index targets, the whole unstated group can be
back-solved.

```{r dose-backsolve}
pub_dose   <- c(1.10, 1.66, 2.28)
pub_factor <- c(2.34, 3.53, 4.86)

# Dose / factor is constant if the equation was applied consistently. The
# published doses carry only three significant figures, so score the spread
# RELATIVE to its own mean rather than absolutely; the three ratios agree to
# 0.24%, which is what rounding alone would produce.
ratio <- pub_dose / pub_factor
stopifnot(diff(range(ratio)) / mean(ratio) < 0.005)

# ratio = MIC * CL_day / (fu * F)  =>  CL_day / fu = ratio * F / MIC
cl_day_over_fu <- mean(ratio) * fbio / mic

implied <- tibble::tibble(
  `Clearance used` = c("IV CL (Table 1)", "IM CL/F (Table 1)"),
  `CL (L/h/kg)`    = c(0.09, 0.15),
  `Implied fu`     = c(0.09, 0.15) * 24 / cl_day_over_fu
)

knitr::kable(implied, digits = 3,
             caption = "Unbound fraction implied by the three published doses, for each reported clearance.")

cat(sprintf("Back-solved CL(per day) / fu = %.3f L/day/kg, i.e. CL/fu = %.4f L/h/kg\n",
            cl_day_over_fu, cl_day_over_fu / 24))
```

Neither reported clearance reproduces the published doses at a physically
possible unbound fraction: the intramuscular `CL/F` of 0.15 L/h/kg requires an
`fu` of `r sprintf("%.3f", 0.15 * 24 / cl_day_over_fu)`, which exceeds 1, and
the intravenous `CL` of 0.09 L/h/kg requires an `fu` of
`r sprintf("%.3f", 0.09 * 24 / cl_day_over_fu)`. Applying the same equation with
the model's own parameters and `fu = 1` gives systematically lower doses:

```{r dose-model}
dose_cmp <- tibble::tibble(
  Endpoint = c("Bacteriostatic", "Bactericidal", "Bacterial elimination"),
  `Index target` = pub_factor,
  `Dose, model (mg/kg)` = pub_factor * mic * (cl_pub * 24) / fbio,
  `Dose, Gao 2025 (mg/kg)` = pub_dose
) |>
  dplyr::mutate(Ratio = `Dose, Gao 2025 (mg/kg)` / `Dose, model (mg/kg)`)

knitr::kable(dose_cmp, digits = 3,
             caption = "Dose equation evaluated with the model's IV clearance and fu = 1, versus the published doses.")
```

The Part 3 Monte Carlo independently favours the lower column. A target
attainment rate of 46.38% at 1 mg/kg means the label dose sits essentially *on*
the bactericidal target for a median-MIC isolate, which implies a bactericidal
dose near 1 mg/kg -- close to the
`r sprintf("%.2f", 3.53 * mic * (cl_pub * 24) / fbio)` mg/kg the equation gives
here, and not the 1.66 mg/kg the paper reports. The discrepancy is a factor of
`r sprintf("%.2f", mean(dose_cmp$Ratio))` across all three endpoints, consistent
with a single unstated `fu` (or an inconsistent clearance) rather than an
arithmetic slip. This vignette does not tune anything to close the gap; see the
Errata.

# Assumptions and deviations

- **No compartmental PK model was published.** Gao 2025 reports
  non-compartmental results only (Section 2.4, Table 1). The one-compartment
  structure with first-order intramuscular absorption is imposed here as the
  smallest structure the published statistics determine. It is not the authors'
  model, because the authors did not fit one.
- **`vc` and `ka` are derived, not printed.** Both come from Table 1 values via
  standard one-compartment identities, and both are corroborated by independent
  published numbers (`vc` by the 12 h concentration in Figure 2, `ka` by the
  reported `Cmax`). They carry inline provenance comments in the model file.
- **The intramuscular terminal half-life is not reproduced.** The model gives
  2.35 h against the published 4.16 h. The flip-flop parameterisation that
  matches 4.16 h is refuted by `Tmax` (4.45 h predicted versus 2.16 h
  published) and `Cmax` (0.387 versus 0.89 ug/mL), and by the parallel terminal
  slopes in Figure 2.
- **The distribution phase is not modelled.** Figure 2 shows a biexponential
  intravenous decline, and `MRT0-last` (2.67 h) is shorter than `1/kel`
  (3.39 h), but a two-compartment model needs four disposition parameters and
  only three independent intravenous summaries are published.
- **Serum MIC, not broth MIC.** `mic` is fixed at the foal-serum value of
  0.062 ug/mL rather than the Mueller Hinton broth value of 0.125 ug/mL,
  following the paper's own argument (Discussion) that broth MICs are
  artificially elevated -- 1.8-fold on average across the 20 isolates of
  Table 2.
- **Emax and E0 are remapped, not changed.** Gao 2025 Section 2.9 defines its
  `Emax` as the no-drug change and its `E0` as the drug-treated change, the
  reverse of the usual convention. The packaged model carries 2.88 as canonical
  `e0` and -4.65 as canonical `emax`; both numbers are unaltered.
- **No variability of any kind is encoded.** Gao 2025 fitted no population
  model. The `+/-` values in Table 1 are standard deviations of per-foal
  non-compartmental estimates, not estimated variance components, so there are
  no `eta` terms and `addSd` is fixed at zero. The Part 3 Monte Carlo therefore
  builds its log-normal AUC distribution explicitly in the vignette from the
  Table 1 mean and SD, exactly as the paper describes, rather than drawing it
  from the model.
- **Post-antibiotic effects are not modelled.** Gao 2025 Table 3 reports PAEs of
  0.13-0.58 h over 1x to 4x MIC. The published PK/PD integration does not
  include a PAE term and neither does this model.
- **Ex vivo exposure convention.** In Part 2 the ex vivo index for a serum
  sample drawn at time `t` is taken as `C(t) x 24 h / (MIC x 24 h)` =
  `C(t) / MIC`, the standard convention for a fixed-concentration incubation.
  Gao 2025 does not print the per-sample index values it used.

# Errata

- **The bacterial-elimination target is internally inconsistent.** Gao 2025
  Table 4 gives 4.86 for bacterial elimination, but its own sigmoid parameters
  put a 4 log10 reduction at `r sprintf("%.3f", index_for(-4))`; an index of
  4.86 corresponds to `r sprintf("%.3f", effect_of(4.86))` log10 CFU/mL. The
  bacteriostatic (2.34) and bactericidal (3.53) targets both reproduce to within
  0.5%, so the transcription of Emax, E0, EC50 and N is sound and the
  elimination entry is the outlier. The packaged model carries the four fitted
  parameters, so a user who back-solves will get
  `r sprintf("%.3f", index_for(-4))` rather than 4.86.
- **The dose equation cannot be evaluated as printed.** Section 2.11 requires an
  unbound fraction `fu` that the paper never reports. Back-solving from the
  three published doses requires an `fu` above 1 if the intramuscular `CL/F` is
  used, or about `r sprintf("%.2f", 0.09 * 24 / cl_day_over_fu)` if the
  intravenous `CL` is used. The published doses are a factor of
  `r sprintf("%.2f", mean(dose_cmp$Ratio))` higher than the same equation
  evaluated with the model's parameters at `fu = 1`, and the paper's own Monte
  Carlo target attainment rates favour the lower values.
- **`CL` and `AUC0-last` in Table 1 are not mutually consistent.** `Dose / CL`
  gives 11.1 h\*ug/mL intravenously against a reported `AUC0-last` of 12.33, and
  the intramuscular `CL` of 0.15 L/h/kg is not consistent with the intravenous
  `CL` of 0.09 L/h/kg and `F` of 43.86% (which imply `CL/F` = 0.205 L/h/kg).
  These are separately averaged per-foal estimates, so the gaps are the expected
  mean-of-ratios versus ratio-of-means artefact rather than errors.
- **Table 1 in the published PDF is split across a page break** and the "Cont."
  block repeats the `Cmax` and `Tmax` rows. The values used here are taken from
  the complete table on page 7.
- No erratum or corrigendum for this article was found on the MDPI article
  landing page as of the extraction date.
