---
title: "BAY81 8973 (Garmann 2017)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BAY81 8973 (Garmann 2017)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Garmann D, McLeay S, Shah A, Vis P, Maas Enriquez M, Ploeger BA.
  Population pharmacokinetic characterization of BAY 81-8973, a full-length
  recombinant factor VIII: lessons learned -- importance of including samples
  with factor VIII levels below the quantitation limit. *Haemophilia.*
  2017 Jul;23(4):528-537.
  doi:[10.1111/hae.13192](https://doi.org/10.1111/hae.13192)
* Description: Two-compartment population PK model for BAY 81-8973 (Kovaltry,
  full-length unmodified recombinant human factor VIII) in patients with severe
  haemophilia A aged 1-61 years pooled from the LEOPOLD I, II, and Kids
  trials.
* Article: <https://doi.org/10.1111/hae.13192>
* Modality: Recombinant factor VIII concentrate, IV infusion (typical infusion
  duration 10 min).

BAY 81-8973 (octocog alfa) is an unmodified full-length recombinant human
factor VIII produced in baby hamster kidney cells and intended for replacement
therapy in haemophilia A. Garmann 2017 developed a two-compartment population
PK model for FVIII activity (chromogenic assay) on data pooled from the three
LEOPOLD trials (LEOPOLD I and II in adults / adolescents 12-61 years; LEOPOLD
Kids part A in subjects <= 12 years). The chosen final model is fit using the
NONMEM M3 likelihood method, which incorporates samples below the lower limit
of quantitation (BLQ) rather than excluding them; the central message of the
paper is that excluding BLQ samples leads to systematically over-estimated
terminal half-lives and therefore under-estimated time spent below a given
FVIII threshold. The packaged model uses the **final M3 parameter estimates**
(Table 2 of the paper, identical to the right-hand "final model" column of
Table 3).

The structural model is a two-compartment IV model with lean body weight (LBW)
as the only covariate, applied as a power function on both clearance and
central volume of distribution; the reference LBW is 51.1 kg, the cohort
median.

# Population

The development cohort comprised **183 male patients with severe haemophilia
A** pooled from the three LEOPOLD trials (Garmann 2017 Table 1):

- All severe haemophilia A (FVIII activity < 1% by one-stage clotting assay).
- Age 1-61 years (median 22, mean 23.3, SD 14.6).
- Body weight 11-124 kg (median 60, mean 57.7, SD 24.8).
- Height 74-192 cm (median 170, mean 160, SD 25.8).
- BMI 13-38.3 kg/m^2 (median 20.4, mean 21.3, SD 5.14).
- Lean body weight 9.25-79.2 kg (median 51.1, mean 46.2, SD 16.7); used as
  the covariate-centering reference.
- Race: 132 White, 31 Asian, 10 Black, 9 Hispanic, 1 Other.
- All male (haemophilia A is X-linked).
- Dosing: per-protocol prophylactic IV BAY 81-8973; the analysis included
  dense PK profiles (10 samples in adults / 4 samples in children following a
  single infusion) from 43 patients plus pre-dose and 30-min recovery samples
  from all 183 patients.
- 1535 chromogenic FVIII activity observations were used in the analysis;
  16.5% of samples were below the lower limit of quantitation (1.5 IU/dL for
  the majority; 3 IU/dL for a small number).

The same metadata is available programmatically via
`readModelDb("Garmann_2017_BAY81_8973")$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in
`inst/modeldb/specificDrugs/Garmann_2017_BAY81_8973.R`. The table below
collects them in one place for review.

| Parameter (model name)         | Value          | Source                                                                |
|--------------------------------|----------------|-----------------------------------------------------------------------|
| `lcl` (typical CL, dL/h)       | log(1.88)      | Garmann 2017 Table 2: CL = 1.88 dL/h                                  |
| `lvc` (typical Vc, dL)         | log(30.0)      | Garmann 2017 Table 2: Vc = 30.0 dL                                    |
| `lq`  (typical CLp, dL/h)      | log(1.90)      | Garmann 2017 Table 2: CLp = 1.90 dL/h                                 |
| `lvp` (typical Vp, dL)         | log(6.37)      | Garmann 2017 Table 2: Vp = 6.37 dL                                    |
| `e_lbw_cl` (LBW power on CL)   | 0.610          | Garmann 2017 Table 2: effect of LBW on CL = 0.610 (95% CI 0.45-0.75)  |
| `e_lbw_vc` (LBW power on Vc)   | 0.950          | Garmann 2017 Table 2: effect of LBW on Vc = 0.950 (95% CI 0.89-1.02)  |
| Reference LBW                  | 51.1 kg        | Garmann 2017 Table 2 footnote: parameterised as `theta * (LBW/51.1)^theta_xx` |
| `etalcl` (omega^2)             | 0.12826        | Garmann 2017 Table 2: BSV CL = 37.0% CV; log-normal omega^2 = log(1 + 0.370^2) |
| `etalvc` (omega^2)             | 0.01247        | Garmann 2017 Table 2: BSV Vc = 11.2% CV; log-normal omega^2 = log(1 + 0.112^2) |
| `propSd`                       | 0.267          | Garmann 2017 Table 2: proportional RUV = 26.7% CV                     |
| `addSd`                        | 1.10 IU/dL     | Garmann 2017 Table 2: additive RUV = 1.10 IU/dL                       |
| Equation: `d/dt(central)`      | n/a            | Garmann 2017 Methods & Figure 2: two-compartment IV model             |
| Equation: `d/dt(peripheral1)`  | n/a            | Garmann 2017 Methods & Figure 2                                       |
| Combined RUV form              | `C * (1 + e1) + e2` | Garmann 2017 Methods, Eq. for `C_ij`                             |
| Reference half-lives           | t_half_alpha = 1.85 h, t_half_beta = 13.9 h | Garmann 2017 Table 2 footnote                    |

The IIV variances above were computed from the reported CV% values via the
log-normal identity `omega^2 = log(1 + (CV/100)^2)`, consistent with the BSV
model `P_j = P_pop * exp(eta_j)` stated in Garmann 2017 Methods. Garmann 2017
tested a CL:Vc covariance during model development but did not retain it
(deltaOBJ = 4.8, below the prespecified retention threshold of 7.9), so
`etalcl` and `etalvc` are independent. No IIV was reported on CLp (Q) or Vp.

# Virtual cohort

Original observed FVIII activity data are not publicly available. The
simulations below use a virtual cohort whose demographics approximate the
Garmann 2017 development population, stratified into the four age bands used
in the paper's PK summary (Table 4): 0 - < 6 years, 6 - < 12 years,
12 - < 18 years, and >= 18 years. Body weight is sampled from a log-normal
distribution centered on a typical weight-for-age within each band; lean body
weight (LBW) is then derived from a male-specific approximation of the James
(1976) formula:
`LBW = 1.10 * WT - 128 * (WT / HT)^2` with WT in kg and HT in m, capped at the
observed Garmann 2017 LBW range (9.25 - 79.2 kg). This formula is used here
purely to populate the covariate column for simulation; Garmann 2017 itself
does not specify the body-composition formula it used to derive LBW.

```{r cohort}
set.seed(2017)

james_lbw_male <- function(wt_kg, ht_m) {
  pmin(pmax(1.10 * wt_kg - 128 * (wt_kg / (ht_m * 100))^2, 9.25), 79.2)
}

make_cohort <- function(n, age_min, age_max,
                        wt_log_mean, wt_log_sd, wt_lo, wt_hi,
                        ht_log_mean, ht_log_sd, ht_lo, ht_hi,
                        label, id_offset = 0L) {
  wt <- pmin(pmax(rlnorm(n, log(wt_log_mean), wt_log_sd), wt_lo), wt_hi)
  ht <- pmin(pmax(rlnorm(n, log(ht_log_mean), ht_log_sd), ht_lo), ht_hi)
  tibble(
    ID        = id_offset + seq_len(n),
    AGE       = runif(n, age_min, age_max),
    WT        = wt,
    HT        = ht,
    LBW       = james_lbw_male(wt, ht),
    age_group = label
  )
}

# Per-band sample sizes mirror Garmann 2017 Table 4 (n=24, 27, 23, 109).
cohort <- bind_rows(
  make_cohort( 24, 1,  6,
              wt_log_mean = 16, wt_log_sd = 0.18, wt_lo = 11, wt_hi = 25,
              ht_log_mean = 1.05, ht_log_sd = 0.07, ht_lo = 0.74, ht_hi = 1.20,
              label = "0-<6 y",      id_offset =   0L),
  make_cohort( 27, 6, 12,
              wt_log_mean = 30, wt_log_sd = 0.18, wt_lo = 18, wt_hi = 50,
              ht_log_mean = 1.30, ht_log_sd = 0.05, ht_lo = 1.10, ht_hi = 1.55,
              label = "6-<12 y",     id_offset = 100L),
  make_cohort( 23, 12, 18,
              wt_log_mean = 55, wt_log_sd = 0.18, wt_lo = 35, wt_hi = 90,
              ht_log_mean = 1.65, ht_log_sd = 0.05, ht_lo = 1.45, ht_hi = 1.85,
              label = "12-<18 y",    id_offset = 200L),
  make_cohort(109, 18, 61,
              wt_log_mean = 75, wt_log_sd = 0.20, wt_lo = 50, wt_hi = 124,
              ht_log_mean = 1.75, ht_log_sd = 0.04, ht_lo = 1.55, ht_hi = 1.92,
              label = ">=18 y",      id_offset = 300L)
) |>
  mutate(age_group = factor(age_group,
                            levels = c("0-<6 y", "6-<12 y",
                                       "12-<18 y", ">=18 y")))

stopifnot(!anyDuplicated(cohort$ID))
summary(cohort)
```

A single 50 IU/kg IV bolus dose of BAY 81-8973 is administered at time 0 and
FVIII activity is observed over a 96-h window.

```{r events}
obs_grid <- c(0, 0.083, 0.25, 0.5, 1, 2, 4, 6, 8, 12,
              seq(18, 96, by = 6))

build_events <- function(pop) {
  # Garmann 2017 uses LBW; nlmixr2lib's canonical column is LBM (same quantity).
  pop <- pop |> mutate(LBM = LBW)
  amt_per_subject <- pop$WT * 50
  d_dose <- pop |>
    mutate(AMT = amt_per_subject) |>
    tidyr::crossing(TIME = 0) |>
    mutate(EVID = 1, CMT = "central", DV = NA_real_)
  d_obs <- pop |>
    tidyr::crossing(TIME = obs_grid) |>
    mutate(AMT = NA_real_, EVID = 0, CMT = "central", DV = NA_real_)
  bind_rows(d_dose, d_obs) |>
    arrange(ID, TIME, desc(EVID)) |>
    as.data.frame()
}

events <- build_events(cohort)
```

# Simulation

Run a stochastic VPC-style simulation (between-subject variability on CL and
Vc included) and a typical-value simulation with the etas zeroed for direct
parameter back-checks.

```{r simulate}
mod <- readModelDb("Garmann_2017_BAY81_8973")

sim <- rxode2::rxSolve(mod, events = events,
                       keep = c("age_group", "WT", "LBM"),
                       returnType = "data.frame")
sim <- sim[sim$time >= 0, ]

mod_typ <- rxode2::zeroRe(mod)
sim_typ <- rxode2::rxSolve(mod_typ, events = events,
                           keep = c("age_group"),
                           returnType = "data.frame")
sim_typ <- sim_typ[sim_typ$time >= 0, ]
```

# FVIII activity-time profiles by age group

Garmann 2017 Figure 5 shows the visual predictive check of the final model
across the development cohort. The figure below reproduces the typical-value
median + 5 - 95% prediction interval of FVIII activity vs. time after dose,
stratified by the four age bands used in the paper's per-band PK summary
(Table 4). The terminal slope is shallower in older / heavier patients
because LBW power-scaling on Vc (exponent 0.950) is steeper than the LBW
power-scaling on CL (exponent 0.610), so the apparent first-order elimination
rate constant of the central compartment, `CL / Vc`, scales as
`LBW^(0.610 - 0.950) = LBW^(-0.34)` and is higher in smaller patients.

```{r profile-plot, fig.width = 8, fig.height = 5}
sim_summary <- sim |>
  filter(time > 0) |>
  group_by(time, age_group) |>
  summarise(
    median = stats::median(Cc, na.rm = TRUE),
    lo     = stats::quantile(Cc, 0.05, na.rm = TRUE),
    hi     = stats::quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(sim_summary, aes(time, median, colour = age_group, fill = age_group)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.18, colour = NA) +
  geom_line(linewidth = 1) +
  scale_y_log10() +
  geom_hline(yintercept = 1.5, linetype = "dashed", colour = "grey40") +
  annotate("text", x = max(obs_grid), y = 1.7,
           label = "LLOQ = 1.5 IU/dL", hjust = 1, vjust = 0,
           colour = "grey30", size = 3) +
  labs(
    x        = "Time after dose (h)",
    y        = "FVIII activity (IU/dL, log scale)",
    colour   = "Age group",
    fill     = "Age group",
    title    = "Simulated BAY 81-8973 PK after a single 50 IU/kg IV dose",
    subtitle = "Stochastic median + 5-95% prediction interval; analogue of Garmann 2017 Figure 5",
    caption  = "Model: Garmann 2017 Haemophilia 23(4):528-537"
  ) +
  theme_bw()
```

# Typical CL and Vss back-check

Garmann 2017 Table 2 reports, for the reference patient (LBW = 51.1 kg),
CL = 1.88 dL/h, Vc = 30.0 dL, CLp = 1.90 dL/h, and Vp = 6.37 dL. Table 2
also reports a typical distribution half-life t_half_alpha = 1.85 h and a
typical elimination half-life t_half_beta = 13.9 h. Reproducing those numbers
from the typical-value simulation is the simplest possible self-consistency
check.

```{r typical-cl-vss}
ev_ref <- rxode2::et(amt = 50 * 70, time = 0, cmt = "central") |>
  rxode2::et(seq(0, 0))
sim_ref <- rxode2::rxSolve(
  mod_typ, events = ev_ref,
  params = data.frame(LBM = 51.1),
  returnType = "data.frame"
)

ref_pars <- sim_ref[1, c("cl", "vc", "q", "vp"), drop = FALSE]
ref_pars$Vss <- ref_pars$vc + ref_pars$vp

# Hybrid rate constants and biphasic half-lives derived from CL, Vc, Q, Vp.
kel <- ref_pars$cl / ref_pars$vc
k12 <- ref_pars$q  / ref_pars$vc
k21 <- ref_pars$q  / ref_pars$vp
trace_term <- kel + k12 + k21
disc       <- sqrt(trace_term^2 - 4 * kel * k21)
alpha_h    <- 0.5 * (trace_term + disc)
beta_h     <- 0.5 * (trace_term - disc)
ref_pars$thalf_alpha <- log(2) / alpha_h
ref_pars$thalf_beta  <- log(2) / beta_h

knitr::kable(
  ref_pars,
  caption = "Typical-value PK parameters at the reference 51.1 kg LBW",
  digits  = c(3, 2, 3, 2, 2, 2, 2)
)
```

The model returns CL = `r round(ref_pars$cl, 2)` dL/h, Vc = `r round(ref_pars$vc, 1)` dL,
CLp = `r round(ref_pars$q, 2)` dL/h, Vp = `r round(ref_pars$vp, 2)` dL, and
Vss = Vc + Vp = `r round(ref_pars$Vss, 1)` dL with biphasic half-lives
t_half_alpha = `r round(ref_pars$thalf_alpha, 2)` h and
t_half_beta = `r round(ref_pars$thalf_beta, 1)` h -- matching the values
reported in Garmann 2017 Table 2 (CL = 1.88 dL/h, Vc = 30.0 dL,
CLp = 1.90 dL/h, Vp = 6.37 dL; t_half_alpha = 1.85 h, t_half_beta = 13.9 h).

# PKNCA validation

Use PKNCA to compute Cmax, AUC0-inf and terminal half-life by age group, and
compare the simulated AUC, weight-normalised CL and weight-normalised Vss
against Garmann 2017 Table 4, which reports geometric mean (geometric % CV)
PK parameters from empirical Bayes individual estimates after a 50 IU/kg
dose.

```{r pknca}
sim_nca <- sim |>
  filter(!is.na(Cc), Cc > 0) |>
  select(id, time, Cc, age_group)

dose_df <- events |>
  filter(EVID == 1) |>
  transmute(id = ID, time = TIME, amt = AMT, age_group, WT)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | age_group + id,
                             concu = "IU/dL",
                             timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | age_group + id,
                             doseu = "IU")

intervals <- data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE,
  clast.obs  = TRUE,
  lambda.z   = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
nca_tbl <- as.data.frame(nca_res$result)

geo_mean    <- function(x) exp(mean(log(x[x > 0]), na.rm = TRUE))
geo_cv_pct  <- function(x) {
  s <- stats::sd(log(x[x > 0]), na.rm = TRUE)
  100 * sqrt(exp(s^2) - 1)
}

auc_summary <- nca_tbl |>
  filter(PPTESTCD == "aucinf.obs") |>
  group_by(age_group) |>
  summarise(
    n             = sum(!is.na(PPORRES)),
    auc_geo_mean  = geo_mean(PPORRES),
    auc_geo_cv    = geo_cv_pct(PPORRES),
    .groups       = "drop"
  )

# Combine with WT and per-subject Vc, Vp from the simulation to derive
# CL/kg = dose/AUC/WT and Vss/kg = (Vc + Vp)/WT for comparison with Table 4.
per_subject <- sim |>
  group_by(id, age_group) |>
  summarise(
    WT  = first(WT),
    LBM = first(LBM),
    Vss = first(vc) + first(vp),
    .groups = "drop"
  )

auc_per_subject <- nca_tbl |>
  filter(PPTESTCD == "aucinf.obs") |>
  transmute(id, age_group, AUC = PPORRES)

combined <- per_subject |>
  left_join(auc_per_subject, by = c("id", "age_group")) |>
  mutate(
    CL_per_kg  = (50 * WT) / AUC / WT,
    Vss_per_kg = Vss / WT
  )

cl_vss_summary <- combined |>
  group_by(age_group) |>
  summarise(
    n              = sum(!is.na(AUC)),
    auc_geo_mean   = geo_mean(AUC),
    auc_geo_cv     = geo_cv_pct(AUC),
    cl_geo_mean    = geo_mean(CL_per_kg),
    cl_geo_cv      = geo_cv_pct(CL_per_kg),
    vss_geo_mean   = geo_mean(Vss_per_kg),
    vss_geo_cv     = geo_cv_pct(Vss_per_kg),
    .groups        = "drop"
  )

knitr::kable(
  cl_vss_summary,
  caption = paste0("Simulated geometric-mean PK parameters by age group ",
                   "after a single 50 IU/kg IV BAY 81-8973 dose."),
  digits  = c(0, 0, 0, 1, 3, 1, 2, 1)
)
```

## Comparison against Garmann 2017 Table 4

Garmann 2017 Table 4 reports geometric mean (geometric % CV) AUC, CL/kg and
Vss/kg in four age bands derived from empirical Bayes individual estimates
across all LEOPOLD patients. Differences > 20% from the published geometric
means would indicate a coding problem.

```{r table4-comparison}
published <- tibble::tribble(
  ~age_group,       ~published_auc,  ~published_auc_cv,
  ~published_cl,    ~published_cl_cv,
  ~published_vss,   ~published_vss_cv,
  "0-<6 y",         970,    25,    0.05,  25,    0.92,  11,
  "6-<12 y",        1242,   35,    0.04,  35,    0.77,  15,
  "12-<18 y",       1523,   27,    0.03,  27,    0.61,  14,
  ">=18 y",         1858,   38,    0.03,  38,    0.56,  14
) |>
  mutate(age_group = factor(age_group,
                            levels = c("0-<6 y", "6-<12 y",
                                       "12-<18 y", ">=18 y")))

comparison <- published |>
  left_join(cl_vss_summary |>
              select(age_group,
                     simulated_auc = auc_geo_mean,
                     simulated_cl  = cl_geo_mean,
                     simulated_vss = vss_geo_mean),
            by = "age_group") |>
  mutate(
    auc_pct_diff = round(100 * (simulated_auc - published_auc) /
                           published_auc, 1),
    cl_pct_diff  = round(100 * (simulated_cl  - published_cl)  /
                           published_cl, 1),
    vss_pct_diff = round(100 * (simulated_vss - published_vss) /
                           published_vss, 1)
  )

knitr::kable(
  comparison,
  caption = paste0("Simulated vs. Garmann 2017 Table 4 PK parameters ",
                   "by age group at 50 IU/kg dose."),
  digits  = c(0, 0, 3, 0, 2, 0, 0, 3, 2, 1, 1, 1)
)
```

# Errata

No published erratum was located for Garmann 2017 (Haemophilia
2017;23(4):528-537). The packaged parameter values are taken from Garmann 2017
Table 2 (which is identical to the right-hand "final model" column of
Table 3) and the equations stated in the Methods section. The reference LBW
of 51.1 kg matches the cohort median LBW reported in Table 1 (51.1 kg,
range 9.25-79.2; n = 182).

# Assumptions and deviations

- **Lean body weight (LBW) covariate stored as canonical `LBM`.** Garmann 2017
  uses the column name `LBW` (lean body weight); the canonical column in
  `inst/references/covariate-columns.md` is `LBM` (lean body mass). LBW and
  LBM refer to the same biological quantity (total body weight minus body
  fat); only the abbreviation differs by literature convention. The
  `covariateData` entry in the model file records `source_name = "LBW"`.
- **LBW body-composition formula not specified.** Garmann 2017 does not state
  the formula used to compute LBW from anthropometric measurements. In
  haemophilia popPK literature LBW is most commonly computed via the Hume
  (1966) or James (1976) formula (which give similar results); the
  Janmahasatian (2005) FFM formula (the canonical `FFM` column) is also
  sometimes used. The vignette's virtual cohort uses a male-specific James
  approximation `LBW = 1.10 * WT - 128 * (WT/HT)^2` purely to populate the
  covariate column for simulation; it does not prescribe a specific formula
  for downstream users.
- **No CL:Vc covariance in IIV.** Garmann 2017 tested a CL:Vc covariance term
  during covariate analysis and reported `deltaOBJ = 4.8`, below the
  prespecified 7.9 retention threshold; the covariance was therefore excluded
  from the final model. The packaged model carries independent `etalcl` and
  `etalvc`.
- **No IIV on CLp (Q) or Vp.** Garmann 2017 Table 2 reports BSV only on CL
  (37.0% CV) and Vc (11.2% CV); no BSV is estimated for the
  intercompartmental clearance or the peripheral volume.
- **Combined RUV interpreted as `C * (1 + e1) + e2`.** Garmann 2017 Methods
  states `C_ij = C_hat * (1 + eps_prop) + eps_add`, with `propSd` the SD of
  the proportional error and `addSd` the SD of the additive error. The
  packaged model uses `Cc ~ add(addSd) + prop(propSd)`, which implements
  exactly this form in nlmixr2.
- **Asian race tested but not retained.** Asian race was the only other
  significant covariate in the forward selection step (effect on Vc with
  `deltaOBJ >= 6.63`, P < 0.01), but its removal in backward elimination
  produced an `OBJ` increase of 6.68 points (below the 7.9 threshold), so it
  was not retained. The packaged model has no race covariate, consistent with
  the published final model.
- **Infusion duration handled at the event-table level.** BAY 81-8973 is given
  as a short IV infusion. Garmann 2017 fixed the infusion duration to
  0.167 h (10 min) where unknown from dosing records. The packaged model is
  agnostic to infusion duration; users supply an `RATE` (or duration) column
  in their event table to model the infusion shape if needed. The vignette's
  simulations use an instantaneous bolus, which is a reasonable approximation
  for a 10-min IV infusion when judging trough activity but slightly
  overestimates very-early Cmax (within ~0.2 h of dose end).
- **No baseline FVIII offset.** The chromogenic assay measures total FVIII
  activity, but in severe haemophilia A baseline endogenous FVIII activity is
  by definition < 1 IU/dL (the inclusion criterion). The packaged model
  treats simulated `Cc` as the activity attributable to BAY 81-8973 above
  this near-zero baseline; an additive baseline (e.g., 0.5 IU/dL) can be
  added downstream by the user if desired.
- **Virtual cohort.** Per-band sample sizes (n = 24, 27, 23, 109) match
  Garmann 2017 Table 4. Body weight and height within each band were
  simulated from log-normal distributions tuned to span the observed
  Garmann 2017 Table 1 ranges; LBW was derived from a male James (1976)
  approximation. Joint covariate structure (e.g., narrow weight range within
  each pediatric age year, or actual within-trial age-weight correlations) is
  not simulated.
