---
title: "Apixaban (Gaspar 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Apixaban (Gaspar 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Gaspar_2023_apixaban"))$reference`
* Description: `r rxode2::rxode(readModelDb("Gaspar_2023_apixaban"))$description`
* Article: <https://doi.org/10.1002/psp4.13032>

Gaspar 2023 is, in the authors' words, the first population PK model of
apixaban built on data collected in a *real-world* population of hospitalized
patients rather than in a clinical trial. Its practical purpose is
pre-procedural: to predict how long after the last dose a patient's apixaban
concentration falls below the 50 ng/mL threshold that current consensus regards
as safe for an invasive procedure, and to show that the standardized
interruption schedules in the PAUSE study are too short for patients with
impaired renal function or reduced P-glycoprotein activity.

# Population

The model was fitted to 825 apixaban concentrations from 100 hospitalized
patients enrolled in the OptimAT study (NCT03477331), a single-centre
prospective observational study at the Geneva University Hospitals running from
January 2018 to November 2019 (Gaspar 2023 Methods, "Study population"). Each
participant underwent one day of multiple finger-prick dried-blood-spot
sampling after their morning dose, scheduled predose and at 0.5, 1, 2, 3, 4, 6
and 8 h, giving a median of 8 samples per patient (range 7-8). Six patients were
hospitalized twice, so the 100 patients contributed 106 occasions.

Per Table 1, the cohort is elderly and renally impaired: median age 77 years
(range 51-94), median body weight 75 kg (range 44-126), 42% female, and a median
Cockcroft-Gault creatinine clearance of only 57.2 mL/min (range 23-136).
Apixaban was prescribed for atrial fibrillation in 89% of patients and for
venous thromboembolism in 11%, at 2.5 mg b.i.d. (n = 40), 5 mg b.i.d. (n = 56)
or 10 mg b.i.d. (n = 4). Phenotyping with the Geneva cocktail classified 47% of
the cohort as P-glycoprotein poor metabolizers, 51% as normal and 2% as
ultra-rapid; the corresponding CYP3A4/5 split was 19% / 78% / 3%. An independent
cohort of 63 patients from the DAPHNE study (NCT03112525) was used for external
validation, where the model showed a mean prediction error of 0% (95% CI -1% to
1%) and a precision of 11%.

Apixaban was quantified in whole blood by LC-MS/MS and converted to a plasma
concentration with the Foerster relationship
`Conc_plasma = Conc_DBS * 1.46 + 13.84`, so the model predicts *plasma*
concentrations.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Gaspar_2023_apixaban")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Gaspar_2023_apixaban.R`. The table below
collects them in one place for review. All values come from Table 2, column
"Final model Estimate (RSE %)", except where noted.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `ltlag` (Tlag) | 0.16 h | Table 2 (RSE 19%; bootstrap 0.18 [0.07-0.28]) |
| `lka` (ka) | 0.82 1/h | Table 2 (RSE 11%; bootstrap 0.92 [0.54-1.87]) |
| `lcl` (CL/F) | 3.2 L/h | Table 2 (RSE 6%; bootstrap 3.1 [2.4-3.6]) |
| `lvc` (Vc/F) | 27 L | Table 2 (RSE 8%; bootstrap 29 [18-44]) |
| `lq` (Q/F) | 16 L/h | Table 2 (RSE 12%; bootstrap 14 [8.0-27]) |
| `lvp` (Vp/F) | 51 L | Table 2 (RSE 12%; bootstrap 56 [25-118]) |
| `e_crcl_cl` (beta_CLcr) | 0.47 | Table 2 (RSE 19%; bootstrap 0.47 [0.22-0.74]) |
| `e_auc_fexo_cl` (beta_P-gp) | -0.17 | Table 2 (RSE 32%; bootstrap -0.21 [-0.44 to -0.05]) |
| CRCL centring constant | 100 mL/min | Methods "Covariate model" and the Table 2 note equation |
| AUC_FEXO centring constant | 205 mg*h/L | Table 2 note equation |
| `etaltlag` | 108% CV -> 0.77307 | Table 2 "IIV Tlag (CV%)"; omega^2 = log(CV^2 + 1) |
| `etalka` | 58% CV -> 0.28998 | Table 2 "IIV ka (CV%)" |
| `etalq` | 65% CV -> 0.35242 | Table 2 "IIV Q (CV%)" |
| `etalvp` | 88% CV -> 0.57346 | Table 2 "IIV Vp (CV%)" |
| `etaiov_cl_1/2` | 30% CV -> 0.08618 | Table 2 "IOV CL (CV%)"; one occasion per hospitalization (Methods) |
| `propSd` | 0.084 | Table 2 "Proportional error model (CV%)" = 8.4% |
| Covariate model form `(Cov/Cov_weight)^beta * exp(eta)` | n/a | Methods "Covariate model", Equation 1 |
| Final clearance equation | n/a | Table 2 note: `CL_occ,i = CL * (CLcr/100)^beta_CLcr * (AUC2-6/205)^beta_PgP * exp(IOV_CL,occ,i)` |
| Two-compartment structure, first-order absorption + lag | n/a | Results "Base model" |
| No IIV on CL (replaced by IOV); no IIV on Vc | n/a | Results "Base model" (dOFV = -110 for IOV; dOFV = 1.7, p > 0.05 for IIV on Vc) |
| Proportional residual error | n/a | Results "Base model" |
| Phenotype-category P-gp AUC2-6 means | 285.5 / 100.1 / 50.4 mg*h/L | Methods "Phenotyping" and Table 3 footnote |
| CKD stage CLcr bands | see below | Methods "Model-based simulations" |
| 50 ng/mL pre-procedural target | 50 ng/mL | Methods "Model-based simulations"; Table 3 |
| 230 ng/mL bleeding-risk threshold | 230 ng/mL | Methods "Model-based simulations"; Figure 2 |

# Deterministic checks against the paper's own derived quantities

Before simulating a cohort, three quantities that follow *deterministically*
from the Table 2 parameters can be checked against numbers the paper prints
elsewhere. These are pure algebra on the transcribed values, so they are the
sharpest available test that the transcription is right, and they are gated
tightly.

```{r deterministic}
ui <- rxode2::rxode(readModelDb("Gaspar_2023_apixaban"))
th <- setNames(ui$theta, names(ui$theta))

cl_typ <- exp(th[["lcl"]])
vc     <- exp(th[["lvc"]])
q      <- exp(th[["lq"]])
vp     <- exp(th[["lvp"]])

# Typical clearance for the reference patient the paper's Results section uses:
# a P-gp NORMAL metabolizer (AUC2-6 = 100.1 mg*h/L) with normal renal function
# (CLcr = 100 mL/min). Note that this is NOT the bare 3.2 L/h of Table 2: the
# equation centres the P-gp term at 205 mg*h/L, not at the NM mean.
cl_nm <- cl_typ * (100 / 100)^th[["e_crcl_cl"]] *
  (100.1 / 205)^th[["e_auc_fexo_cl"]]

# Two-compartment macro rate constants.
twocmt_halflives <- function(cl, vc, q, vp) {
  k10 <- cl / vc
  k12 <- q / vc
  k21 <- q / vp
  a <- k10 + k12 + k21
  b <- k10 * k21
  lam1 <- (a + sqrt(a^2 - 4 * b)) / 2
  lam2 <- (a - sqrt(a^2 - 4 * b)) / 2
  c(alpha = log(2) / lam1, beta = log(2) / lam2)
}
hl <- twocmt_halflives(cl_nm, vc, q, vp)

# Covariate effect sizes relative to a P-gp normal metabolizer.
pgp_ratio <- function(auc) (auc / 205)^th[["e_auc_fexo_cl"]] / (100.1 / 205)^th[["e_auc_fexo_cl"]]

det <- tibble::tibble(
  Quantity = c(
    "Typical CL/F, P-gp NM, CLcr 100 (L/h)",
    "Distribution half-life t1/2 alpha (h)",
    "Terminal half-life t1/2 beta (h)",
    "Total volume of distribution Vc + Vp (L)",
    "CL/F ratio, P-gp PM vs NM",
    "CL/F ratio, P-gp UM vs NM"
  ),
  Model = c(cl_nm, hl[["alpha"]], hl[["beta"]], vc + vp,
            pgp_ratio(285.5), pgp_ratio(50.4)),
  Published = c(NA, 0.70, 17, 78, 0.84, 1.12),
  `Source` = c(
    "derived",
    "Results 'Covariate analyses': 0.70 h (alpha)",
    "Results 'Covariate analyses': 17 h (beta)",
    "Discussion: 'The total volume of distribution of 78 L'",
    "Results: CL 'reduced by 16%' in P-gp PM",
    "Results: CL 'increased by 12%' in P-gp UM"
  )
)
knitr::kable(det, digits = 3,
             caption = "Quantities derived algebraically from the Table 2 parameters, against the values Gaspar 2023 reports in prose.")

stopifnot(
  # Deterministic algebra on transcribed constants -- no simulation, no cohort,
  # so these are gated to the paper's own reporting precision. A mis-transcribed
  # volume, clearance or exponent moves each of them by tens of percent.
  abs(hl[["alpha"]] - 0.70) < 0.05,
  abs(hl[["beta"]] - 17) < 1.5,
  abs((vc + vp) - 78) < 0.5,
  abs(pgp_ratio(285.5) - 0.84) < 0.01,
  abs(pgp_ratio(50.4) - 1.12) < 0.01
)
```

The half-lives, the total volume of distribution and both P-glycoprotein effect
sizes reproduce to the precision the paper reports them. This also confirms the
reading of the centring constant 205 mg\*h/L in the Table 2 note equation: had
`CL = 3.2 L/h` instead been the clearance of a P-gp *normal* metabolizer, the
terminal half-life would come out at 18.4 h rather than the 16.5 h that rounds
to the published 17 h, and the alpha half-life would be unchanged, so the beta
half-life is what discriminates the two readings.

One published derived quantity is **not** reproduced; see
"Assumptions and deviations" below.

# Virtual cohort

Original observed data are not publicly available. The cohort below follows the
paper's own simulation design (Methods, "Model-based simulations"): 5 chronic
kidney disease stages crossed with 3 P-glycoprotein phenotypes, with a uniform
distribution assumed over each covariate's range.

```{r cohort}
# set.seed() seeds R's RNG. It does NOT seed rxode2's simulation RNG, and
# rxode2's streams are partitioned PER SOLVER THREAD, so this cohort is
# reproducible on this machine and different on a machine with a different
# thread count. Every assertion below is written to hold for ANY cohort the
# model can produce; see pattern 12 of the known-vignette-failure-patterns
# reference.
set.seed(20231001)

# 50 per arm across 15 arms (750 subjects). Sized against the render budget:
# the vignette must stay well inside the 5-minute ceiling because the
# consolidation merge re-renders every vignette in parallel, where CPU
# contention multiplies wall-clock time. Measured single-threaded here at
# 40 / 75 / 100 per arm: 54 / 168 / 203 s (the PKNCA step scales worse than
# linearly), so this sits at roughly 80 s with ample headroom. The only gates
# that depend on cohort size are the Table 3 medians, which are checked on
# robust statistics rather than extremes.
n_per_arm <- 50L

# CKD bands per Methods "Model-based simulations". Stage 1 is defined as
# CLcr > 90 mL/min with no upper bound, so it is capped at the observed cohort
# maximum of 136 mL/min (Table 1). Stage 5 is defined as CLcr < 15 mL/min and
# is capped below at 5 mL/min; both stage 4 and stage 5 extend below the
# observed cohort minimum of 23 mL/min and are therefore extrapolation, which
# the paper also performs.
ckd_bands <- tibble::tribble(
  ~ckd,      ~crcl_lo, ~crcl_hi,
  "Stage 1",       90,      136,
  "Stage 2",       60,       89,
  "Stage 3",       30,       59,
  "Stage 4",       15,       29,
  "Stage 5",        5,       15
)

# P-gp phenotype bands. Methods "Phenotyping" gives category mean +/- SD for the
# fexofenadine AUC(2-6 h) index; Methods "Model-based simulations" says a
# uniform distribution over the studied range was assumed, which is read here as
# uniform over mean +/- SD. See "Assumptions and deviations".
pgp_bands <- tibble::tribble(
  ~pgp,  ~auc_mean, ~auc_sd,
  "PM",      285.5,    67.1,
  "NM",      100.1,    47.5,
  "UM",       50.4,    15.3
) |>
  mutate(auc_lo = auc_mean - auc_sd, auc_hi = auc_mean + auc_sd)

arms <- tidyr::crossing(ckd_bands, pgp_bands) |>
  mutate(arm = paste(ckd, pgp)) |>
  arrange(ckd, factor(pgp, levels = c("PM", "NM", "UM")))

# Steady-state 5 mg b.i.d., delivered as a single steady-state dose record
# (`ss = 1`, `ii = 12`) at t = 0 rather than as a 20-dose burn-in. Time 0 is
# therefore the last drug intake, which is exactly the time origin Gaspar 2023
# Figure 3 and Table 3 use. This was verified against an explicit 20-dose q12h
# solve: the two agree to a relative 7e-5, and it removes 228 h of burn-in
# integration per subject from the render.
#
# The model is strictly linear, so the 2.5 and 10 mg columns of Table 3 are
# recovered later by rescaling the concentration threshold rather than by
# re-simulating.
tau <- 12

# Observation grid, graded: fine over the dosing interval (for Cmin / Cmax /
# AUC at steady state), then progressively coarser out to 240 h, where Table 3
# censors. The published medians span 6-149 h, so the grid resolution is kept
# well under 5% of the crossing time throughout.
obs_times <- sort(unique(c(
  seq(0, 12, by = 0.25),
  seq(12, 48, by = 1),
  seq(48, 120, by = 3),
  seq(120, 240, by = 6)
)))

make_arm <- function(n, arm, crcl_lo, crcl_hi, auc_lo, auc_hi, id_offset) {
  subj <- tibble::tibble(
    id       = id_offset + seq_len(n),
    arm      = arm,
    CRCL     = runif(n, crcl_lo, crcl_hi),
    AUC_FEXO = runif(n, auc_lo, auc_hi),
    # One hospitalization per simulated patient, so the first IOV eta applies.
    OCC      = 1L
  )
  doses <- subj |>
    mutate(time = 0, amt = 5, evid = 1L, cmt = "depot", ss = 1L, ii = tau)
  obs <- subj |>
    tidyr::crossing(time = obs_times) |>
    # Observations are placed on the ODE state `central`, never on the
    # algebraic observable `Cc`; rxode2 returns Cc as a column regardless.
    mutate(amt = NA_real_, evid = 0L, cmt = "central", ss = 0L, ii = 0)
  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}

events <- do.call(bind_rows, lapply(seq_len(nrow(arms)), function(i) {
  a <- arms[i, ]
  make_arm(n_per_arm, a$arm, a$crcl_lo, a$crcl_hi, a$auc_lo, a$auc_hi,
           id_offset = (i - 1L) * n_per_arm)
}))

# Disjoint IDs across arms are mandatory: rxSolve treats id as the subject key
# and would silently merge duplicates into one over-dosed subject.
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
stopifnot(nrow(dplyr::distinct(events, id)) == n_per_arm * nrow(arms))
```

# Simulation

```{r simulate}
mod <- readModelDb("Gaspar_2023_apixaban")
sim <- rxode2::rxSolve(mod, events = events,
                       keep = c("arm", "CRCL", "AUC_FEXO")) |>
  as.data.frame()

# `Cc` is the individual prediction; it carries no residual error (the
# proportional error term is applied at the observation level, not here), which
# is what the paper's simulations of Cmin and stopping time use.
sim <- sim |>
  mutate(
    ckd  = sub("^(Stage \\d).*$", "\\1", arm),
    pgp  = sub("^Stage \\d ", "", arm),
    # Time 0 is the last drug intake, so time after dose is the time column.
    tad  = time
  )
stopifnot(all(!is.na(sim$Cc)), all(sim$Cc >= 0))
```

# Replicate Figure 2: steady-state Cmin by CKD stage and P-gp activity

Figure 2 of Gaspar 2023 shows the distribution of the simulated steady-state
minimum concentration for each CKD stage and P-gp phenotype, against the 5th-95th
percentile band of Cmin reported by the manufacturer in atrial fibrillation
(41-230 ng/mL for 5 mg b.i.d.) and the 230 ng/mL concentration associated with a
20% increased bleeding risk.

```{r figure-2}
# Replicates Figure 2 of Gaspar 2023 (5 mg b.i.d. panel).
cmin_tbl <- sim |>
  filter(tad >= 0, tad <= 12) |>
  group_by(id, arm, ckd, pgp) |>
  summarise(cmin = min(Cc), .groups = "drop") |>
  mutate(pgp = factor(pgp, levels = c("PM", "NM", "UM")))

ggplot(cmin_tbl, aes(x = ckd, y = cmin, fill = pgp)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 41, ymax = 230,
           alpha = 0.15, fill = "grey40") +
  geom_hline(yintercept = 230, linewidth = 0.6) +
  geom_boxplot(outlier.size = 0.5, position = position_dodge(width = 0.8)) +
  scale_y_log10() +
  labs(x = "Chronic kidney disease stage", y = "Steady-state Cmin (ng/mL)",
       fill = "P-gp activity",
       title = "Figure 2 - steady-state Cmin, apixaban 5 mg b.i.d.",
       caption = paste("Replicates Figure 2 of Gaspar 2023. Grey band: 41-230 ng/mL,",
                       "the manufacturer's 5th-95th percentile of Cmin in atrial",
                       "fibrillation. Black line: 230 ng/mL bleeding-risk threshold."))
```

The paper makes three quantitative claims about this figure. Checking them on
the medians of the cohort above would be checking a statistic with meaningful
sampling noise at 50 subjects per arm, so they are instead evaluated on a
**typical-value (variability-suppressed) solve** at each stratum's midpoint
covariates. That makes each number a deterministic property of the transcribed
parameters, and the gates correspondingly sharp.

```{r figure-2-claims}
ckd_mid <- ckd_bands |> mutate(crcl = (crcl_lo + crcl_hi) / 2) |> select(ckd, crcl)
# Phenotype-category MEANS from Methods "Phenotyping", not the band midpoints.
pgp_mid <- pgp_bands |> select(pgp, auc = auc_mean)

grid <- tidyr::crossing(ckd_mid, pgp_mid) |> mutate(id = dplyr::row_number())

grid_ev <- bind_rows(
  grid |> mutate(time = 0, amt = 5, evid = 1L, cmt = "depot", ss = 1L, ii = tau),
  grid |> tidyr::crossing(time = seq(0, tau, by = 0.05)) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central", ss = 0L, ii = 0)
) |>
  mutate(CRCL = crcl, AUC_FEXO = auc, OCC = 1L) |>
  # Drop the lower-case staging columns before the solve: carrying both `crcl`
  # and `CRCL` through rxSolve collides on a case-insensitive name match.
  select(id, time, amt, evid, cmt, ss, ii, CRCL, AUC_FEXO, OCC, ckd, pgp) |>
  arrange(id, time, desc(evid))

typ_cmin <- rxode2::rxSolve(mod |> rxode2::zeroRe(), events = grid_ev,
                            keep = c("ckd", "pgp")) |>
  as.data.frame() |>
  filter(!is.na(Cc)) |>
  group_by(ckd, pgp) |>
  summarise(cmin = min(Cc), .groups = "drop")

ref  <- typ_cmin |> filter(ckd == "Stage 1") |> select(pgp, ref = cmin)
rise <- typ_cmin |> left_join(ref, by = "pgp") |>
  mutate(pct_vs_stage1 = 100 * (cmin / ref - 1))

pm_um <- typ_cmin |>
  tidyr::pivot_wider(names_from = pgp, values_from = cmin) |>
  mutate(pct_pm_vs_um = 100 * (PM / UM - 1))

knitr::kable(
  rise |>
    # Drop `ref` before pivoting: it varies by phenotype, so leaving it in
    # makes pivot_wider treat each (stage, phenotype) as its own row and the
    # table renders one value per row with NAs everywhere else.
    select(-ref) |>
    tidyr::pivot_wider(names_from = pgp, values_from = c(cmin, pct_vs_stage1)) |>
    select(ckd, cmin_PM, cmin_NM, cmin_UM,
           pct_vs_stage1_PM, pct_vs_stage1_NM, pct_vs_stage1_UM) |>
    dplyr::rename(
      "CKD stage"        = ckd,
      "Cmin PM (ng/mL)"  = cmin_PM,
      "Cmin NM (ng/mL)"  = cmin_NM,
      "Cmin UM (ng/mL)"  = cmin_UM,
      "% vs stage 1, PM" = pct_vs_stage1_PM,
      "% vs stage 1, NM" = pct_vs_stage1_NM,
      "% vs stage 1, UM" = pct_vs_stage1_UM
    ),
  digits = 0,
  caption = "Typical-value steady-state Cmin by CKD stage and P-gp phenotype (5 mg b.i.d.), and the increase over stage 1 within each phenotype."
)

knitr::kable(
  pm_um |> select(ckd, pct_pm_vs_um) |>
    dplyr::rename("CKD stage" = ckd, "PM vs UM Cmin difference (%)" = pct_pm_vs_um),
  digits = 0,
  caption = "Difference in typical-value Cmin between P-gp poor and ultra-rapid metabolizers. The paper states 20-35%; see Assumptions and deviations."
)

claim_s3 <- rise |> filter(ckd == "Stage 3") |> pull(pct_vs_stage1)

stopifnot(
  # Claim: "patients with stage 3 CKD would have a median Cmin 65%-80% higher
  # than patients with a normal renal function, independently of their P-gp
  # activities". Deterministic solve, so this is gated close to the paper's own
  # interval -- widened only for the stage-1 and stage-3 midpoint choice.
  all(claim_s3 > 60), all(claim_s3 < 90),
  # "independently of their P-gp activities": the stage-3 rise must be nearly
  # the same for all three phenotypes.
  diff(range(claim_s3)) < 12,
  # Structural orderings, deterministic and large: exposure rises monotonically
  # with CKD stage for every phenotype, and a poor metabolizer is always more
  # exposed than an ultra-rapid one.
  all(rise$pct_vs_stage1[rise$ckd == "Stage 2"] < rise$pct_vs_stage1[rise$ckd == "Stage 3"]),
  all(rise$pct_vs_stage1[rise$ckd == "Stage 3"] < rise$pct_vs_stage1[rise$ckd == "Stage 4"]),
  all(rise$pct_vs_stage1[rise$ckd == "Stage 4"] < rise$pct_vs_stage1[rise$ckd == "Stage 5"]),
  all(pm_um$PM > pm_um$NM), all(pm_um$NM > pm_um$UM)
)
```

The stage-3 claim reproduces closely and, as the paper says, almost
independently of P-gp phenotype. The stage-5 magnitude and the P-gp
poor-versus-ultra-rapid difference do not reproduce inside the paper's stated
ranges; both are analysed in "Assumptions and deviations" and are deliberately
excluded from the gate rather than having the gate widened around them.

# Replicate Table 3: pre-operative time to reach 50 ng/mL

Table 3 reports the time after the last dose at which apixaban falls below 50
ng/mL, by dose, CKD stage and P-gp activity, censored at 240 h.

The model is strictly linear in dose: bioavailability is absorbed into the
apparent parameters, there is no saturable elimination and no dose covariate.
The whole 36-cell table can therefore be recovered from the single 5 mg b.i.d.
simulation above by rescaling the *threshold* rather than re-simulating, because
crossing 50 ng/mL on a dose D is exactly crossing `50 * 5 / D` ng/mL on the 5 mg
profile. The linearity is asserted rather than assumed.

```{r linearity-check}
# Verify dose-proportionality before relying on it: solve one small arm at
# 2.5 mg and confirm the profile is exactly half the 5 mg profile.
lin_ev <- events |> filter(id <= 5L)
lin_5  <- rxode2::rxSolve(mod |> rxode2::zeroRe(), events = lin_ev) |> as.data.frame()
lin_25 <- rxode2::rxSolve(mod |> rxode2::zeroRe(),
                          events = lin_ev |> mutate(amt = ifelse(evid == 1L, 2.5, amt))) |>
  as.data.frame()
stopifnot(nrow(lin_5) == nrow(lin_25))
# Exact linearity: relative deviation is solver tolerance, not physiology.
stopifnot(max(abs(lin_25$Cc * 2 - lin_5$Cc) / pmax(lin_5$Cc, 1e-8)) < 1e-4)
```

```{r table-3}
# First time after the last dose at which the concentration has dropped below
# `thr` and stays below it, on the hourly grid; censored at 240 h as Table 3 is.
time_below <- function(tad, cc, thr, cap = 240) {
  above <- which(cc >= thr)
  if (length(above) == 0L) return(0)
  t_last_above <- tad[max(above)]
  nxt <- tad[tad > t_last_above]
  if (length(nxt) == 0L) return(cap)
  min(nxt[1], cap)
}

wash <- sim |>
  filter(tad >= 0) |>
  arrange(id, tad)

stop_times <- wash |>
  group_by(id, ckd, pgp) |>
  summarise(
    `2.5mg` = time_below(tad, Cc, 50 * 5 / 2.5),
    `5mg`   = time_below(tad, Cc, 50 * 5 / 5),
    `10mg`  = time_below(tad, Cc, 50 * 5 / 10),
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(c(`2.5mg`, `5mg`, `10mg`),
                      names_to = "dose", values_to = "t_stop")

sim_tab3 <- stop_times |>
  group_by(ckd, pgp, dose) |>
  summarise(sim_med = median(t_stop), .groups = "drop")

# Gaspar 2023 Table 3, median column (h). Table 3 covers CKD stages 1-4 only.
published_tab3 <- tibble::tribble(
  ~ckd,      ~pgp,  ~`2.5mg`, ~`5mg`, ~`10mg`,
  "Stage 1", "PM",        11,     30,      50,
  "Stage 1", "NM",         7,     22,      39,
  "Stage 1", "UM",         6,     17,      32,
  "Stage 2", "PM",        17,     40,      64,
  "Stage 2", "NM",        12,     32,      53,
  "Stage 2", "UM",         9,     27,      46,
  "Stage 3", "PM",        32,     62,      91,
  "Stage 3", "NM",        22,     46,      72,
  "Stage 3", "UM",        16,     37,      60,
  "Stage 4", "PM",        64,    107,     149,
  "Stage 4", "NM",        44,     77,     111,
  "Stage 4", "UM",        36,     69,     100
) |>
  tidyr::pivot_longer(c(`2.5mg`, `5mg`, `10mg`),
                      names_to = "dose", values_to = "pub_med")

cmp3 <- published_tab3 |>
  left_join(sim_tab3, by = c("ckd", "pgp", "dose")) |>
  mutate(pct_diff = 100 * (sim_med - pub_med) / pub_med)
stopifnot(nrow(cmp3) == 36L, !anyNA(cmp3$sim_med))

knitr::kable(
  cmp3 |>
    mutate(dose = factor(dose, levels = c("2.5mg", "5mg", "10mg"))) |>
    arrange(ckd, factor(pgp, levels = c("PM", "NM", "UM")), dose) |>
    dplyr::rename(
      "CKD stage"        = ckd,
      "P-gp"             = pgp,
      "Regimen (b.i.d.)" = dose,
      "Published (h)"    = pub_med,
      "Simulated (h)"    = sim_med,
      "Difference (%)"   = pct_diff
    ),
  digits = 1,
  caption = "Median time after the last dose to fall below 50 ng/mL: Gaspar 2023 Table 3 against the packaged model."
)

stopifnot(
  # Centre of the distribution across all 36 cells. A mis-transcribed clearance,
  # volume, exponent or centring constant shifts every cell in the same
  # direction and blows this immediately.
  abs(median(cmp3$pct_diff)) < 15,
  # Envelope, robust to which subjects land in the tails of a 100-subject
  # median. Deliberately NOT a max(): the extreme of a random cohort is not
  # reproducible across rxode2 builds or solver thread counts.
  quantile(abs(cmp3$pct_diff), 0.9) < 35,
  # The published ordering is a large, structural effect (a factor of ~3.5 from
  # the shortest to the longest cell), so it must survive any cohort.
  cmp3$sim_med[cmp3$ckd == "Stage 4" & cmp3$pgp == "PM" & cmp3$dose == "10mg"] >
    cmp3$sim_med[cmp3$ckd == "Stage 1" & cmp3$pgp == "UM" & cmp3$dose == "2.5mg"]
)
```

# Replicate Figure 3: washout profiles after the last dose

```{r figure-3}
# Replicates Figure 3 of Gaspar 2023: (a) P-gp normal metabolizer with stage 1
# CKD, (b) P-gp poor metabolizer with stage 4 CKD, both on 5 mg b.i.d.
fig3 <- sim |>
  filter(tad >= 0, tad <= 120,
         (ckd == "Stage 1" & pgp == "NM") | (ckd == "Stage 4" & pgp == "PM")) |>
  mutate(panel = ifelse(ckd == "Stage 1",
                        "(a) stage 1 CKD, P-gp NM",
                        "(b) stage 4 CKD, P-gp PM")) |>
  group_by(panel, tad) |>
  summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = median(Cc),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot(fig3, aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line() +
  geom_hline(yintercept = 50, colour = "red") +
  facet_wrap(~panel) +
  labs(x = "Time after last apixaban intake (h)", y = "Apixaban (ng/mL)",
       title = "Figure 3 - washout at steady state, 5 mg b.i.d.",
       caption = paste("Replicates Figure 3 of Gaspar 2023. Median and 90% prediction",
                       "interval. Red line: the 50 ng/mL pre-procedural target."))
```

# PKNCA validation

NCA is run over the final steady-state dosing interval (228-240 h) for each of
the 15 arms.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc), time <= tau) |>
  dplyr::select(id, time, Cc, arm)

# Guarantee a time = 0 record per (id, arm) so PKNCA can anchor AUC.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, arm, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

dose_df <- events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, arm)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)

intervals <- data.frame(
  start = 0, end = 12,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_summary <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "cmin", "auclast")) |>
  dplyr::group_by(arm, PPTESTCD) |>
  dplyr::summarise(median = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median)

knitr::kable(
  nca_summary |>
    dplyr::rename(
      "Arm"                     = arm,
      "Cmax (ng/mL)"            = cmax,
      "Tmax (h)"                = tmax,
      "Cmin (ng/mL)"            = cmin,
      "AUC0-12 (ng*h/mL)"       = auclast
    ),
  digits = 1,
  caption = "Median steady-state NCA parameters over the 12 h dosing interval, apixaban 5 mg b.i.d."
)
```

## Comparison against published NCA

Gaspar 2023 publishes no NCA table of its own. The two NCA-comparable numbers
it does report are the terminal half-life of the reference patient and the
manufacturer's Cmin percentile band, so the comparison below uses a
typical-value (variability-suppressed) solve, where the half-life is a
deterministic property of the model rather than a tmax-selected, NA-prone
statistic over a cohort with 88% CV on the peripheral volume.

```{r nca-typical}
typ_subj <- tibble::tibble(id = 1L, CRCL = 100, AUC_FEXO = 100.1, OCC = 1L)
typ_events <- dplyr::bind_rows(
  typ_subj |> dplyr::mutate(time = 0, amt = 5, evid = 1L, cmt = "depot",
                            ss = 1L, ii = tau),
  typ_subj |> tidyr::crossing(time = seq(0, 240, by = 0.5)) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", ss = 0L, ii = 0)
) |>
  dplyr::arrange(time, dplyr::desc(evid))

sim_typ <- rxode2::rxSolve(mod |> rxode2::zeroRe(), events = typ_events) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc))

# Terminal slope fitted well after the washout transient, per pattern 11 of the
# known-vignette-failure-patterns reference: a time-to-50% measured from the
# moment of withdrawal includes the distribution phase and reads long.
term <- sim_typ |> dplyr::filter(time >= 48, time <= 240, Cc > 0)
hl_fit <- log(2) / -coef(lm(log(Cc) ~ time, data = term))[["time"]]

published_nca <- tibble::tibble(
  arm = "Typical patient (CLcr 100 mL/min, P-gp NM)",
  half.life = 17
)
sim_nca_typ <- tibble::tibble(
  arm = "Typical patient (CLcr 100 mL/min, P-gp NM)",
  half.life = hl_fit
)

knitr::kable(
  sim_nca_typ |>
    dplyr::left_join(published_nca, by = "arm", suffix = c("_sim", "_pub")) |>
    dplyr::mutate(`Difference (%)` = 100 * (half.life_sim - half.life_pub) / half.life_pub) |>
    dplyr::rename(
      "NCA parameter"                     = arm,
      "t1/2 simulated (h)"                = half.life_sim,
      "t1/2 published (h)"                = half.life_pub
    ),
  digits = 2,
  caption = "Terminal half-life of the typical patient against the value reported in Gaspar 2023 Results."
)

stopifnot(
  # Deterministic (zeroRe) solve of a linear model, so this is a numerical
  # identity check against the analytic 16.5 h, not a cohort statistic.
  abs(hl_fit - hl[["beta"]]) < 0.5,
  abs(hl_fit - 17) < 1.5
)
```

# Assumptions and deviations

- **The paper's "45% lower CL" in stage 4 CKD is not reproduced by its own
  printed equation, and the equation was followed.** Results, "Covariate
  analyses" states that a typical P-gp NM patient with stage 4 CKD (mean
  CLcr = 22 mL/min) has "45% lower CL" than the same patient with stage 1 CKD
  (mean CLcr = 100 mL/min). The Table 2 note equation with the Table 2 estimate
  `beta_CLcr = 0.47` gives `(22/100)^0.47 = 0.491`, i.e. **50.9% lower**, and no
  reading of the simulation design recovers 45% either (integrating the power
  term over uniform CLcr on the stage-4 and stage-1 bands gives 53.8% lower).
  Recovering exactly 45% would require `beta_CLcr = 0.395`. The abstract's
  companion phrasing -- "a patient with stage 4 chronic kidney disease would
  have a 45% higher drug exposure" -- is inconsistent with both the equation
  and with Figure 2, which shows stage 3 already 65-80% higher and stage 5
  200-250% higher, bracketing stage 4 far above 45%. The printed equation and
  the Table 2 point estimates are therefore taken as authoritative and the prose
  percentage is recorded here as an erratum. The same "45% lower CL" claim
  appears in the abstract and in Results; both are affected. Every other derived
  quantity the paper prints -- the alpha and beta half-lives, the 78 L total
  volume, and both P-gp effect sizes -- reproduces exactly, which is what makes
  this one stand out as a slip rather than a transcription error on this side.
- **Terminal half-lives of 33 h (stage 3) and 52 h (stage 4).** Results also
  reports these as terminal half-lives "on average". The analytic typical-value
  terminal half-lives at the stage midpoints (CLcr 44.5 and 22 mL/min, P-gp NM)
  are 23.4 h and 32.0 h. The gap is consistent with these being *means over the
  simulated cohort*, which are inflated by the right skew that log-normal IIV on
  Q and Vp induces on a half-life; the paper does not state how they were
  computed, so they are not gated here.
- **The stage-5 Cmin rise depends entirely on an unstated simulation bound, and
  is not a model discrepancy.** Results reports that the Cmin increase over
  normal renal function "reaches up to 200%-250% in patients suffering from
  stage 5 CKD". Stage 5 is defined only as `CLcr < 15 mL/min`, with no lower
  bound stated anywhere in the paper, and the whole claim turns on that missing
  number: holding everything else fixed, the typical-value rise over stage 1 is
  227% at CLcr = 14, 247% at 12.5, 290% at 10, 353% at 7.5 and 457% at 5
  mL/min. The paper's 200-250% therefore corresponds to sampling stage 5 close
  to its upper edge (CLcr roughly 12.5-14). The cohort here uses 5-15 mL/min and
  lands at ~290%. Because the target is an artefact of an undocumented choice
  rather than of the model, it is reported but not gated.
- **The P-gp poor-versus-ultra-rapid Cmin difference is larger than the paper
  states.** Results reports "A 20%-35% difference in Cmin is expected between
  P-gp PM and UM patients at all dosage regimens and CKD stages combined"; the
  packaged model gives 38-47% across the five CKD stages, deterministically.
  The discrepancy is structural rather than a transcription problem: the
  clearance ratio between the two phenotypes is exactly
  `(285.5/50.4)^-0.17 = 0.745`, i.e. a 34.3% clearance difference, which sits
  squarely inside the paper's quoted 20-35% band, but a steady-state *trough*
  on a 12 h interval responds by more than the clearance ratio, because reduced
  clearance also lengthens the terminal half-life and so raises the trough.
  The most likely reading is that the quoted range describes the clearance
  effect rather than the simulated Cmin effect. Both P-gp effect sizes on
  clearance itself (-16% for PM, +12% for UM versus NM) reproduce exactly, as
  the deterministic-checks section shows, so the covariate transcription is
  confirmed independently of this claim. Reported but not gated.
- **P-glycoprotein covariate distribution.** Methods, "Model-based simulations"
  states only that "a uniform distribution in the studied range was assumed for
  all covariates", without giving a range for the fexofenadine AUC(2-6 h) index
  within each phenotype category. Methods, "Phenotyping" gives category
  mean +/- SD, so the cohort here draws uniformly over mean +/- SD
  (PM 218.4-352.6, NM 52.6-147.6, UM 35.1-65.7 mg\*h/L). This choice affects the
  width of the simulated distributions but not their centres, which are what the
  gates above test.
- **CKD band endpoints.** Stage 1 is defined by an open-ended CLcr > 90 mL/min
  and is capped here at the observed cohort maximum of 136 mL/min (Table 1);
  stage 5 is defined by CLcr < 15 mL/min and is bounded below at 5 mL/min. Both
  stage 4 (15-29) and stage 5 extend below the observed cohort minimum of
  23 mL/min, so those two strata are extrapolation beyond the fitted range -- as
  they are in the paper's own simulations.
- **Occasion.** Each simulated patient is given `OCC = 1`, i.e. a single
  hospitalization, which is the modal case in the source data (94 of 100
  patients). Because the paper estimates one shared inter-occasion variance,
  the second occasion's variance is `fix()`ed equal to the first in the model
  file and a two-occasion patient would simply draw a second, independent
  clearance offset.
- **Bioavailability.** All disposition parameters are apparent (CL/F, Vc/F,
  Q/F, Vp/F), as the paper labels them; F is not identifiable from oral-only
  data. No `f(depot)` term is applied, so absolute volumes and clearances are
  not directly comparable with intravenous apixaban literature values.
- **Dried blood spots.** The fitted concentrations are plasma values back-
  calculated from dried blood spots via `Conc_plasma = Conc_DBS * 1.46 + 13.84`
  (Foerster et al., cited by Gaspar 2023 Methods). That affine transformation
  is a property of the assay, not of the model, and is not re-applied here.
- **Covariates screened but not retained.** Body weight, age, sex, smoking
  status, urea, albumin, AST, ALT, alkaline phosphatase, gamma-glutamyl
  transferase, bilirubin and CYP3A4/5 phenotypic activity were all tested and
  none was retained; the paper publishes no point estimate for any of them.
  They are recorded in the model file's `covariatesDataExcluded` metadata rather
  than in `covariateData`. Note in particular that body weight and age reach
  clearance *through* `CRCL`, since both are inputs to the Cockcroft-Gault
  equation -- the Discussion makes this argument explicitly -- so a user must
  not add a separate allometric weight term on top of this model.
- **No non-paper-derived parameter values.** Every value in `ini()` comes from
  Gaspar 2023 Table 2 or from the constants printed in the Table 2 note and
  Methods. Nothing was digitised from a figure, obtained by correspondence, or
  carried from an upstream model.
