---
title: "Desmopressin oral lyophilisate in growing piglets (Gasthuys 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Desmopressin oral lyophilisate in growing piglets (Gasthuys 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Gasthuys_2018_desmopressin_piglet"))$reference`
* Description: `r rxode2::rxode(readModelDb("Gasthuys_2018_desmopressin_piglet"))$description`
* Article: <https://doi.org/10.3389/fphar.2018.00041>
* Supplementary Table 1 (model development path):
  <https://www.frontiersin.org/articles/10.3389/fphar.2018.00041/full#supplementary-material>

# Population

Thirty-two male Landrace x large white piglets were studied in four age groups
of eight animals each: 8 days, 4 weeks, 7 weeks and 6 months
(Gasthuys 2018, Table 1). Body weight spans nearly two orders of magnitude
across the groups - a median of 2.21 kg at 8 days versus 113.6 kg at
6 months - which is what makes the cohort informative about size-driven
maturation and is why body weight is the covariate the model retains.

Every animal received a single 120 ug desmopressin oral lyophilisate
(Minirin Melt) placed under the tongue, with the snout held closed for 30 s.
Animals were fasted from 1 h before to 1.5 h after dosing. Plasma was sampled
richly through an indwelling jugular catheter at 0, 5, 15, 30 and 60 min and
at 1.5, 2, 3, 4, 6, 8, 10, 12 and 24 h. Desmopressin was measured by
radioimmunoassay with a 4.2 pg/mL limit of quantification; 8% of
concentrations fell below it and were excluded before fitting.

The study is a juvenile-animal model for the pediatric population: desmopressin
is used to treat primary nocturnal enuresis in children, but full pediatric PK
studies taking growth and maturation into account are lacking.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Gasthuys_2018_desmopressin_piglet")()$population`).

```{r population}
pop <- rxode2::rxode(
  readModelDb("Gasthuys_2018_desmopressin_piglet")
)$population
str(pop, max.level = 1)
```

# Structural model

The paper's Figure 3 shows a two-compartment disposition model fed by two
parallel input paths, which the authors call a "dual input function":

* A fraction `Bio` (86%) of the dose is released **zero-order over a duration
  `D1`** into a first depot, representing buccal absorption of the
  lyophilisate, and reaches the central compartment first-order at `Ka1`.
  No lag time was needed, indicating that sublingual absorption starts
  immediately.
* The remaining `1 - Bio` (14%) is assumed to be swallowed. It enters a second
  depot after a **lag time `Tlag`** of 1 h and is absorbed first-order at
  `Ka2`, representing gastro-intestinal absorption of the swallowed fraction.

The dual input is what reproduces the second peak visible in many individual
plasma profiles. Because no intravenous data were collected, absolute oral
bioavailability `F` is not identifiable, so clearance and both volumes are
apparent (`CL/F`, `V1/F`, `V2/F`) and the two dose fractions sum to one.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Gasthuys_2018_desmopressin_piglet.R`. The
table below collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL/F) | 395 L/h | Table 2, `CL/F = theta1 x (BW/10)^theta11 x e^eta1` (SE 31.8, RSE 8.05%) |
| `lvc` (V1/F) | 131 L | Table 2, `V1/F = theta2 x (BW/10)^theta12 x e^eta2` (SE 21.1, RSE 16.1%) |
| `lka` (Ka1) | 0.275 1/h | Table 2, `Ka1 = theta3 x e^eta3` (SE 0.0272, RSE 9.89%) |
| `lq` (Q/F) | 32 L/h | Table 2, `Q/F = theta4` (SE 6.8, RSE 21.3%) |
| `lvp` (V2/F) | 436 L | Table 2, `V2/F = theta5 x e^eta5` (SE 137, RSE 31.4%) |
| `ld1` (D1) | 0.16 h | Table 2, `D1 = theta6 x e^eta6` (SE 0.0473, RSE 29.6%) |
| `lka2` (Ka2) | 0.399 1/h | Table 2, `Ka2 = theta7` (SE 0.0677, RSE 17.0%) |
| `logitfdepot` (Bio) | 86% | Table 2, `Bio = theta8` (SE 0.0488, RSE 5.67%) |
| `ltlag` (Tlag) | 1 h | Table 2, `Tlag = theta9` (SE 0.00196, RSE 0.20%) |
| `e_wt_cl` | 1.03 | Table 2, "Influence of BW on CL" (SE 0.0627, RSE 6.09%) |
| `e_wt_vc` | 0.691 | Table 2, "Influence of BW on V1" (SE 0.135, RSE 19.54%) |
| `etalcl` | 0.175 | Table 2, "IIV CL/F" (shrinkage 1.97%, RSE 13.20%) |
| `etalvc` | 0.641 | Table 2, "IIV V1/F" (shrinkage 25.4%, RSE 20.59%) |
| `etalka` | 0.0903 | Table 2, "IIV Ka1" (shrinkage 16.0%, RSE 18.77%) |
| `etalvp` | 0.634 | Table 2, "IIV V2/F" (shrinkage 61.8%, RSE 15.93%) |
| `etald1` | 0.485 | Table 2, "IIV D1" (shrinkage 47.4%, RSE 26.29%) |
| `etalogitfdepot` | 0.627 | Table 2, "IIV Bio" (shrinkage 42.9%, RSE 31.18%) |
| IIV on Q/F, Ka2, Tlag | 0 (FIX) | Table 2; Supplementary Table 1 run 7 |
| `propSd` | 0.228 | Table 2, `ADD = theta10` = 22.8% CV (SE 0.0291, RSE 12.76%) |
| Dual-input structure, dose split, `D1`, `Tlag` | n/a | Figure 3 (model scheme) and Results, "Pharmacokinetic Analysis" |
| Body-weight centering on 10 kg | n/a | Table 2 caption |
| Reference AUC0-inf and T-half by age group | see below | Table 3 |
| Reference Cmax, 8-day group | ~250 pg/mL | Discussion ("120 ug; Cmax: +- 250 pg/mL") |

# Structural verification

Before simulating a cohort, four deterministic checks confirm that the packaged
model implements the paper's Figure 3 structure. These are closed-form
identities, not cohort statistics, so they are asserted tightly.

```{r structural-setup}
mod <- readModelDb("Gasthuys_2018_desmopressin_piglet")
mod_typ <- rxode2::zeroRe(mod)

dose_ug <- 120

# Typical animal of each age group, at the group's median body weight (Table 1).
groups <- tibble::tibble(
  cohort = factor(
    c("8 days", "4 weeks", "7 weeks", "6 months"),
    levels = c("8 days", "4 weeks", "7 weeks", "6 months")
  ),
  WT = c(2.21, 10.0, 15.0, 113.6)
)

# Closed-form typical parameters, straight from the ini() values.
theta <- tibble::tibble(
  cl = 395 * (groups$WT / 10)^1.03,
  vc = 131 * (groups$WT / 10)^0.691,
  vp = 436,
  q = 32
) |>
  dplyr::mutate(
    kel = cl / vc,
    k12 = q / vc,
    k21 = q / vp,
    # Terminal (beta) hybrid rate constant of the two-compartment disposition.
    beta = 0.5 *
      (
        (kel + k12 + k21) -
          sqrt((kel + k12 + k21)^2 - 4 * kel * k21)
      ),
    thalf_beta = log(2) / beta
  )

build_events <- function(wt, id = 1L, times, f_depot = TRUE, f_depot2 = TRUE) {
  rows <- list()
  if (f_depot) {
    rows <- c(rows, list(tibble::tibble(
      id = id, time = 0, amt = dose_ug, evid = 1L,
      cmt = "depot", rate = -2, dur = NA_real_
    )))
  }
  if (f_depot2) {
    rows <- c(rows, list(tibble::tibble(
      id = id, time = 0, amt = dose_ug, evid = 1L,
      cmt = "depot2", rate = 0, dur = NA_real_
    )))
  }
  rows <- c(rows, list(tibble::tibble(
    id = id, time = times, amt = NA_real_, evid = 0L,
    cmt = "central", rate = 0, dur = NA_real_
  )))
  dplyr::bind_rows(rows) |>
    dplyr::mutate(WT = wt) |>
    dplyr::arrange(time, dplyr::desc(evid)) |>
    dplyr::select(-dur)
}

# A grid fine enough to resolve the 0.16 h zero-order release and long enough
# (96 h, roughly 10 terminal half-lives) that the extrapolated AUC tail is
# negligible.
grid_long <- sort(unique(c(
  seq(0, 6, by = 0.01),
  seq(6, 24, by = 0.1),
  seq(24, 96, by = 1)
)))

# Exposure spans more than two orders of magnitude across the age groups, and
# the closed-form identities below are asserted tightly, so the solver is run
# with tolerances well inside the assertion bounds rather than at the defaults.
solver_tol <- list(atol = 1e-12, rtol = 1e-10)

solve_typical <- function(wt, times, f_depot = TRUE, f_depot2 = TRUE) {
  ev <- build_events(wt, times = times, f_depot = f_depot, f_depot2 = f_depot2)
  out <- rxode2::rxSolve(
    mod_typ, ev,
    atol = solver_tol$atol, rtol = solver_tol$rtol,
    returnType = "data.frame"
  )
  if (is.null(out$id)) out$id <- 1L
  out
}

trapz <- function(x, y) sum(diff(x) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)
```

## 1. Dose recovery: `CL/F x AUC0-inf = Dose`

Because absolute bioavailability is folded into the apparent parameters, the
two dose fractions sum to exactly one and the model must return the whole dose
through clearance. `AUC` is in pg*h/mL while the dose is in ug and clearance in
L/h, so the identity carries the same factor of 1000 the model applies to `Cc`.

```{r gate-dose-recovery}
recovery <- vapply(
  seq_len(nrow(groups)),
  function(i) {
    s <- solve_typical(groups$WT[i], grid_long)
    auc <- trapz(s$time, s$Cc) + utils::tail(s$Cc, 1) / theta$beta[i]
    theta$cl[i] * auc / 1000 / dose_ug
  },
  numeric(1)
)

knitr::kable(
  groups |>
    dplyr::mutate(
      "CL/F (L/h)" = round(theta$cl, 1),
      "Recovered fraction of dose" = round(recovery, 4)
    ),
  caption = "Dose recovery through clearance, typical animal of each age group."
)

# Closed-form identity, limited only by trapezoidal error on a fine grid.
stopifnot(
  all(abs(recovery - 1) < 0.01),
  !anyNA(recovery)
)
```

## 2. The dose split is 86% / 14% between the two depots

An AUC-recovery gate alone cannot see the dose split: it is satisfied by any
`Bio`. Solving each input path on its own and confirming the two AUCs are in
the published 86:14 ratio - and that together they reproduce the combined
profile exactly - is what pins `Bio`.

```{r gate-dose-split}
split_check <- vapply(
  seq_len(nrow(groups)),
  function(i) {
    both <- solve_typical(groups$WT[i], grid_long)
    only1 <- solve_typical(groups$WT[i], grid_long, f_depot2 = FALSE)
    only2 <- solve_typical(groups$WT[i], grid_long, f_depot = FALSE)
    a_both <- trapz(both$time, both$Cc)
    a1 <- trapz(only1$time, only1$Cc)
    a2 <- trapz(only2$time, only2$Cc)
    c(buccal_fraction = a1 / a_both, additivity = (a1 + a2) / a_both)
  },
  numeric(2)
)

knitr::kable(
  groups |>
    dplyr::mutate(
      "Buccal (depot 1) share of AUC" = round(split_check["buccal_fraction", ], 4),
      "Sum of separate paths / combined AUC" = round(split_check["additivity", ], 6)
    ),
  caption = "Dose split between the two absorption paths (Table 2: Bio = 86%)."
)

stopifnot(
  # Bio = 0.86 exactly (Table 2). A mis-set split moves this by whole percent.
  all(abs(split_check["buccal_fraction", ] - 0.86) < 0.002),
  # The two inputs are strictly parallel, so their AUCs are additive.
  all(abs(split_check["additivity", ] - 1) < 1e-6)
)
```

## 3. The zero-order release really lasts `D1` = 0.16 h

Dose records into `depot` carry `rate = -2`, which is what tells rxode2 to use
the modelled `dur(depot)`. Without it rxode2 silently delivers a bolus and the
zero-order release disappears while every AUC stays identical - so this check
watches the depot amount, not the concentration.

```{r gate-zero-order}
d1 <- 0.16
bio <- 0.86
fine <- seq(0, 0.5, by = 0.0005)
s_fine <- solve_typical(10.0, fine)

# During the release the depot fills at a constant rate bio*Dose/D1 minus the
# small amount already absorbed; after D1 no further drug enters.
inflow_rate <- bio * dose_ug / d1
at <- function(tt) stats::approx(s_fine$time, s_fine$depot, tt)$y

zero_order <- tibble::tibble(
  "Time (h)" = c(0.04, 0.08, 0.12, 0.16, 0.20, 0.40),
  "Depot 1 amount (ug)" = round(at(c(0.04, 0.08, 0.12, 0.16, 0.20, 0.40)), 3),
  "Cumulative input (ug)" = round(pmin(inflow_rate * c(0.04, 0.08, 0.12, 0.16, 0.20, 0.40), bio * dose_ug), 3)
)
knitr::kable(zero_order, caption = "Zero-order filling of depot 1 over D1 = 0.16 h.")

# Total drug that ever entered depot 1 = amount still there at D1 plus the
# amount already absorbed by then. It must equal Bio * Dose.
absorbed_by_d1 <- 0.275 * trapz(
  s_fine$time[s_fine$time <= d1],
  s_fine$depot[s_fine$time <= d1]
)
entered <- at(d1) + absorbed_by_d1

# Input has stopped after D1: the depot must be strictly falling from there on.
after <- s_fine[s_fine$time > d1 + 1e-9, ]
stopifnot(
  abs(entered - bio * dose_ug) < 0.05,
  all(diff(after$depot) < 0),
  # And it was still rising before D1 (a bolus would have jumped at t = 0).
  at(0.02) < at(0.14)
)
```

## 4. The swallowed fraction is withheld for exactly `Tlag` = 1 h

```{r gate-lag}
lag_grid <- sort(unique(c(seq(0, 3, by = 0.0005), 1)))
s_lag <- solve_typical(10.0, lag_grid)

d2_at <- function(tt) stats::approx(s_lag$time, s_lag$depot2, tt)$y

lag_tab <- tibble::tibble(
  "Time (h)" = c(0.5, 0.99, 1.00, 1.01, 1.5),
  "Depot 2 amount (ug)" = round(d2_at(c(0.5, 0.99, 1.00, 1.01, 1.5)), 4)
)
knitr::kable(lag_tab, caption = "Depot 2 is empty until Tlag = 1 h, then receives (1 - Bio) x Dose.")

stopifnot(
  # Nothing in depot 2 before the lag.
  all(s_lag$depot2[s_lag$time < 1 - 1e-6] == 0),
  # The full complementary fraction arrives at the lag.
  abs(d2_at(1.0) - (1 - bio) * dose_ug) < 1e-6,
  # ... and decays first-order at Ka2 = 0.399 1/h afterwards.
  abs(d2_at(2.0) / d2_at(1.0) - exp(-0.399)) < 1e-4
)
```

# Virtual cohort

Original observed data are not publicly available. The simulations below use a
virtual population whose body-weight distribution reproduces the per-group mean
+- SD and observed range of Table 1.

```{r cohort}
# set.seed() seeds R's RNG for the covariate draw. It does NOT seed rxode2's
# simulation RNG, whose streams are partitioned per solver thread, so the
# simulated etas differ between a 2-core CI runner and a 16-thread
# workstation. Every assertion below is written to hold for any cohort the
# model can produce.
set.seed(20180131)
rxode2::rxSetSeed(20180131)

n_per_group <- 100

wt_spec <- tibble::tibble(
  cohort = groups$cohort,
  mean = c(2.01, 10.0, 15.8, 112.9), # Table 1, mean +- SD
  sd = c(0.43, 1.69, 1.98, 9.11),
  lo = c(1.54, 7.0, 14.0, 100.0), # Table 1, observed range
  hi = c(2.64, 12.0, 19.0, 124.0)
)

make_cohort <- function(i, id_offset) {
  wt <- pmin(
    pmax(
      stats::rnorm(n_per_group, wt_spec$mean[i], wt_spec$sd[i]),
      wt_spec$lo[i]
    ),
    wt_spec$hi[i]
  )
  subj <- tibble::tibble(
    id = id_offset + seq_len(n_per_group),
    WT = wt,
    cohort = wt_spec$cohort[i]
  )
  obs_times <- sort(unique(c(
    seq(0, 6, by = 0.02),
    seq(6, 24, by = 0.2),
    seq(24, 96, by = 1)
  )))
  dplyr::bind_rows(
    subj |> dplyr::mutate(time = 0, amt = dose_ug, evid = 1L, cmt = "depot", rate = -2),
    subj |> dplyr::mutate(time = 0, amt = dose_ug, evid = 1L, cmt = "depot2", rate = 0),
    subj |>
      tidyr::crossing(time = obs_times) |>
      dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", rate = 0)
  ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  make_cohort(1, 0L),
  make_cohort(2, 1000L),
  make_cohort(3, 2000L),
  make_cohort(4, 3000L)
)

# Disjoint ids across cohorts: duplicate ids are silently merged by rxSolve
# into a single subject receiving the summed dose.
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))

events |>
  dplyr::filter(evid == 1L, cmt == "depot") |>
  dplyr::group_by(cohort) |>
  dplyr::summarise(
    n = dplyr::n(),
    "Mean WT (kg)" = round(mean(WT), 2),
    "SD WT (kg)" = round(stats::sd(WT), 2),
    "Min WT (kg)" = round(min(WT), 2),
    "Max WT (kg)" = round(max(WT), 2),
    .groups = "drop"
  ) |>
  dplyr::rename("Age group" = cohort) |>
  knitr::kable(caption = "Virtual cohort body weight vs. Gasthuys 2018 Table 1.")
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep = c("WT", "cohort"),
  atol = solver_tol$atol,
  rtol = solver_tol$rtol
) |>
  as.data.frame()

sim <- sim |> dplyr::mutate(cohort = factor(cohort, levels = levels(groups$cohort)))

# Far into the tail the 6-month profiles fall many orders of magnitude below
# the peak, where the integrator can return a concentration a few units in the
# last place below zero. PKNCA would then take log() of a negative number and
# return NaN for aucinf.obs.
#
# Asserting `all(Cc >= 0)` is the wrong guard here: it is satisfied or not
# depending on which subjects a given solver-thread count happens to draw (at
# the default tolerances it held at 1, 4 and 16 threads and failed at 2, with a
# minimum of -6e-9 against a 500 pg/mL peak). The tightened tolerances above
# removed it at every thread count tested, but the robust statement is about
# MAGNITUDE: any negative excursion must be numerically negligible relative to
# the scale of the data. A structural error large enough to matter -- a
# negative dose fraction, say -- would be a sizeable fraction of the peak and
# still trips this.
worst_negative <- min(c(0, sim$Cc), na.rm = TRUE)
stopifnot(
  !all(is.na(sim$Cc)),
  abs(worst_negative) < 1e-6 * max(sim$Cc, na.rm = TRUE)
)

# Having established the negatives are noise, clamp them so PKNCA never sees
# one.
sim$Cc <- pmax(sim$Cc, 0)
```

# Replicate published figures

## Figure 2 - mean concentration-time profiles by age group

```{r figure-2}
# Replicates Figure 2 of Gasthuys 2018: mean desmopressin plasma concentration
# vs. time by age group over the first 24 h.
fig2 <- sim |>
  dplyr::filter(time <= 24) |>
  dplyr::group_by(cohort, time) |>
  dplyr::summarise(mean_cc = mean(Cc), .groups = "drop")

ggplot(fig2, aes(time, mean_cc, colour = cohort)) +
  geom_line(linewidth = 0.7) +
  scale_x_continuous(breaks = seq(0, 24, by = 2)) +
  labs(
    x = "Time post administration (h)",
    y = "Plasma desmopressin (pg/mL)",
    colour = "Age group",
    title = "Figure 2 - mean concentration-time profiles",
    caption = "Replicates Figure 2 of Gasthuys 2018 (120 ug oral lyophilisate)."
  ) +
  theme_bw()
```

The 8-day group reaches a mean peak near 250 pg/mL between 1 and 2 h, the
4-week and 7-week groups peak between roughly 45 and 70 pg/mL, and the
6-month group barely rises above the 4.2 pg/mL assay limit - the ordering and
magnitudes the published Figure 2 shows. The Discussion states the 8-day Cmax
explicitly ("120 ug; Cmax: +- 250 pg/mL"), which is the only Cmax the paper
prints as a number.

```{r figure-2-peaks}
peaks <- fig2 |>
  dplyr::group_by(cohort) |>
  dplyr::summarise(
    "Peak of mean profile (pg/mL)" = round(max(mean_cc), 1),
    "Time of peak (h)" = time[which.max(mean_cc)],
    .groups = "drop"
  ) |>
  dplyr::rename("Age group" = cohort)
knitr::kable(peaks, caption = "Peak of the simulated mean profile by age group.")

cmax_8d <- max(fig2$mean_cc[fig2$cohort == "8 days"])

# The paper prints ~250 pg/mL for this group. Rendering at 1, 2, 4, 8 and 16
# solver threads gave 215.5 / 240.6 / 222.9 / 219.3 / 232.2 pg/mL, i.e. at most
# 13.8% below the published figure. The 25% bound sits outside that spread and
# still fails on a mis-transcribed dose, clearance or volume, which move Cmax
# by tens of percent.
stopifnot(
  abs(cmax_8d - 250) / 250 < 0.25,
  # Exposure must fall monotonically with age group, as Figure 2 shows. This
  # is a large, structural ordering, not a near-zero effect.
  all(diff(peaks$`Peak of mean profile (pg/mL)`) < 0)
)
```

## Figure 5 - visual predictive check

```{r figure-5}
# Replicates Figure 5 of Gasthuys 2018: 5th / 50th / 95th percentiles of the
# simulated concentrations against time after dose, on a log scale, pooled
# across age groups exactly as the published VPC is.
vpc <- sim |>
  dplyr::filter(time <= 24) |>
  dplyr::group_by(cohort, time) |>
  dplyr::summarise(
    Q05 = stats::quantile(Cc, 0.05),
    Q50 = stats::quantile(Cc, 0.50),
    Q95 = stats::quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  dplyr::filter(Q05 > 0)

ggplot(vpc, aes(time, Q50, group = cohort)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95, fill = cohort), alpha = 0.25) +
  geom_line(aes(colour = cohort), linewidth = 0.6) +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)",
    y = "Desmopressin concentration (pg/mL)",
    colour = "Age group",
    fill = "Age group",
    title = "Figure 5 - visual predictive check",
    caption = paste(
      "Replicates Figure 5 of Gasthuys 2018: median and 5th-95th percentile",
      "band of the simulated concentrations."
    )
  ) +
  theme_bw()
```

The published VPC pools all four age groups into one panel, which produces the
three well-separated bands visible in the paper's figure; splitting by age
group here makes the same structure explicit.

# PKNCA validation

NCA is run per subject with the age group as the treatment grouping variable so
that results can be compared with the per-group values of Table 3. The
simulation window extends to 96 h - roughly ten terminal half-lives - so that
`AUC0-inf` is determined by the profile rather than by the extrapolated tail.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, cohort)

# Guarantee a time-zero record per subject; for an extravascular dose the
# pre-dose concentration is 0. Without it PKNCA warns on every subject that
# the AUC interval starts before the first measurement.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |>
    dplyr::distinct(id, cohort) |>
    dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, cohort, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | cohort + id)

# One dose record per subject: the animal received a single 120 ug
# lyophilisate. The two event-table dose rows are the model's internal split
# of that one dose, not two administrations.
dose_df <- events |>
  dplyr::filter(evid == 1L, cmt == "depot") |>
  dplyr::select(id, time, amt, cohort)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | cohort + id)

intervals <- data.frame(
  start = 0,
  end = Inf,
  cmax = TRUE,
  tmax = TRUE,
  aucinf.obs = TRUE,
  half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_long <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  dplyr::group_by(cohort, PPTESTCD) |>
  dplyr::summarise(
    mean = mean(PPORRES, na.rm = TRUE),
    median = stats::median(PPORRES, na.rm = TRUE),
    .groups = "drop"
  )

# Descriptive table uses medians (robust to the right tail that a 43% CV on
# clearance produces); the comparison against Table 3 below uses means,
# because Table 3 reports means.
nca_wide <- nca_long |>
  dplyr::select(cohort, PPTESTCD, median) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median)

nca_mean <- nca_long |>
  dplyr::select(cohort, PPTESTCD, mean) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = mean)

stopifnot(
  nrow(nca_wide) == 4L, nrow(nca_mean) == 4L,
  !anyNA(nca_wide$aucinf.obs), !anyNA(nca_mean$aucinf.obs),
  all(is.finite(nca_mean$aucinf.obs))
)

nca_wide |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) signif(x, 3))) |>
  dplyr::rename(
    "Age group" = cohort,
    "Cmax (pg/mL)" = cmax,
    "Tmax (h)" = tmax,
    "AUC0-inf (pg*h/mL)" = aucinf.obs,
    "t1/2 (h)" = half.life
  ) |>
  knitr::kable(caption = "Median simulated NCA parameters by age group.")
```

## Comparison against published NCA

```{r nca-compare}
# Gasthuys 2018 Table 3, mean +- SD of the individual (EBE-derived) secondary
# parameters.
published <- tibble::tibble(
  cohort = groups$cohort,
  aucinf.obs = c(1369, 436, 215, 32.7),
  half.life = c(3.83, 1.40, 0.90, 0.10)
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_mean |> dplyr::select(cohort, aucinf.obs, half.life),
  reference = published,
  by = "cohort",
  units = c(aucinf.obs = "pg*h/mL", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated (mean of 100 virtual piglets per group) vs. Gasthuys 2018",
    "Table 3 (mean +- SD of 8 animals). * differs from reference by >20%."
  )
)
```

**AUC0-inf** is reproduced across a 40-fold exposure range. This is the
quantity the structural model determines: with bioavailability folded into the
apparent parameters, `AUC0-inf = Dose / (CL/F)`, so agreement here confirms
both the typical clearance and the body-weight exponent. Both columns are
means, so the comparison is like for like. The remaining group-to-group scatter
has no systematic direction - the simulated mean sits above the published mean
in the 8-day group and a little below it in the other three - which is what
eight animals per published group and a 43% CV on clearance produce.

**Half-life is a known deviation and is not gated.** The model's terminal
half-life is a property of its disposition parameters alone and is
essentially independent of age group:

```{r half-life-closed-form}
knitr::kable(
  groups |>
    dplyr::mutate(
      "Model terminal t1/2 (h, closed form)" = round(theta$thalf_beta, 2),
      "Gasthuys 2018 Table 3 T-half-el (h)" = published$half.life
    ) |>
    dplyr::rename("Age group" = cohort),
  caption = "Closed-form terminal half-life of the fitted model vs. the paper's reported T-half-el."
)
```

The fitted disposition parameters (`Q/F` = 32 L/h against `V2/F` = 436 L) give
`k21` = 0.073 1/h, so the true terminal phase of this model has a half-life
near 10 h in every group. The paper's `T-half-el` instead ranges from 3.83 h at
8 days to 0.10 h at 6 months - an eighty-fold spread that no set of fixed
disposition parameters can generate. The explanation is in the data rather than
in the model: `T-half-el` was estimated from each animal's own
concentration-time profile, and how much of the terminal phase is visible
depends entirely on how far the profile stays above the 4.2 pg/mL limit of
quantification. The 8-day animals peak near 250 pg/mL and remain measurable to
24 h; the 6-month animals barely clear the limit at all, so their apparent
"terminal" slope is fitted across the peak and reads as 0.10 h. The reported
`T-half-el` is therefore an assay-window artefact, not a model prediction, and
reproducing it would require reproducing the censoring rather than the
pharmacokinetics. It is reported above for completeness and excluded from the
gate.

```{r nca-gate}
auc_ratio <- nca_mean$aucinf.obs[match(published$cohort, nca_mean$cohort)] /
  published$aucinf.obs
stopifnot(!anyNA(auc_ratio), length(auc_ratio) == 4L)

knitr::kable(
  tibble::tibble(
    "Age group" = published$cohort,
    "Simulated / published AUC0-inf" = round(auc_ratio, 3)
  ),
  caption = "Ratio of simulated to published AUC0-inf (mean vs. mean)."
)

# These are cohort statistics, so the bounds come from a thread-count sweep
# rather than from one run. Rendering at 1, 2, 4, 8 and 16 solver threads gave
# a median ratio of 0.859 / 0.859 / 0.925 / 0.888 / 0.916 and per-group ratios
# spanning 0.78 to 1.36. The bounds below sit outside that range -- do not
# tighten them back to what a single render happens to produce.
stopifnot(
  # Centre: a mis-transcribed clearance, dose, unit or body-weight exponent
  # moves all four ratios together and blows this immediately.
  abs(log(stats::median(auc_ratio))) < log(1.35),
  # Envelope: no single group may drift by more than two-fold, even allowing
  # for eight animals per published group and a 43% CV on clearance.
  all(auc_ratio > 0.5 & auc_ratio < 2.0)
)
```

# Assumptions and deviations

- **Inter-individual variability on `Bio` is carried on the logit scale, not
  the exponential scale.** This is the one interpretive decision in the
  extraction and it is worth stating plainly. The Methods say only that IIV on
  "clearance, volume of distribution and the absorption parameters" was
  log-normal and modelled with an exponential-error model; Table 2 reports
  `IIV Bio` = 0.627 without saying what scale it sits on. An exponential eta of
  that size would put `Bio` above 1 for roughly 42% of animals, making the
  complementary fraction `1 - Bio` negative - not a tail event but nearly half
  the population, which a model with a successful covariance step and the
  well-behaved VPC of Figure 5 cannot have had. Table 2's own notation points
  the same way: every parameter with a non-zero IIV is written as
  `theta x e^eta` **except** `Bio`, which is printed as a bare `Bio = theta8`,
  exactly as the three parameters whose IIV was fixed to zero are. A
  multiplicative eta cannot be written in that column's format if it is not
  multiplicative. The logit scale keeps `Bio` in (0, 1) with a median of 0.86
  and a 5th-95th percentile range of about 0.63-0.96, and is the form
  nlmixr2lib uses elsewhere for bounded dose fractions. If the maintainers
  establish from the original control stream that an exponential eta was
  used, this is a one-line change to `logitfdepot` / `etalogitfdepot` in the
  model file.
- **Body weight is the only covariate implemented.** Body surface area, age and
  GFR were screened by the authors but not retained, and no coefficient is
  published for any of them, so they are documented in the model's
  `covariatesDataExcluded` metadata rather than in `covariateData`. GFR
  reached significance on forward inclusion but failed backward elimination
  (Supplementary Table 1 run 16).
- **Concentration units.** Parameter values are used exactly as printed
  (dose in ug, volumes in L, clearances in L/h). The observation applies an
  explicit factor of 1000 to convert ug/L to the pg/mL the paper reports; no
  parameter value is rescaled.
- **Dose records into `depot` must carry `rate = -2`.** That is what makes
  rxode2 honour the modelled `dur(depot) = D1`. A plain bolus silently ignores
  the zero-order release, and because the total absorbed amount is unchanged
  every AUC-based check still passes - which is why the structural section
  above watches the depot amount directly.
- **The two dose records are one administration.** The event table doses
  `depot` and `depot2` with 120 ug each; `f(depot) = Bio` and
  `f(depot2) = 1 - Bio` split the single administered 120 ug lyophilisate
  between them. The PKNCA dose object therefore carries one 120 ug record per
  subject, not two.
- **Body-weight distribution.** Table 1 reports mean +- SD and median [range]
  per age group but not the individual weights. The virtual cohort draws body
  weight from a normal distribution matched to the group mean and SD and
  truncated to the published range.
- **Table 1 vs. the Methods text.** The Methods report the 7-week group as
  13.9 +- 2.74 kg while Table 1 reports 15.8 +- 1.98 kg. Table 1 is used here,
  since it is the table the model's covariate summary refers to. The choice
  shifts the 7-week simulated AUC by about 12%.
- **Published half-life is not reproduced and is not gated**, for the reasons
  set out above. It is an artefact of how much of each animal's terminal phase
  sat above the assay limit of quantification rather than a prediction of the
  fitted disposition parameters.
- **No erratum or corrigendum** was found for this article.
