---
title: "Metoclopramide (Ge 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metoclopramide (Ge 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Ge_2020_metoclopramide"))$reference`
* Description: `r rxode2::rxode(readModelDb("Ge_2020_metoclopramide"))$description`
* Article: <https://doi.org/10.1111/cts.12803> (open access; PMC7719387)

Ge et al. fit a two-compartment model with first-order absorption to 87
opportunistic plasma metoclopramide concentrations from 50 pediatric patients
dosed per standard of care, pooling intravenous and enteral (oral, nasogastric,
nasojejunal, gastrostomy) data. Body weight entered a priori with fixed
allometric exponents; no other covariate was retained. The final model was then
used to simulate steady-state exposure after 0.1 and 0.15 mg/kg orally every
6 hours in virtual patients from term neonates to adolescents (Figures 3 and 4).

# Population

The analysis population (Ge 2020 Table 1, Results "Patient characteristics")
was 50 patients enrolled across US Pediatric Trials Network sites in the POPS
opportunistic-sampling trial (NCT01431326): 20 infants (postnatal age <= 2
years), 9 children (2-12 years) and 21 adolescents (> 12 years), with median
postnatal age 8.89 years (range 0.01-19.13) and median body weight 23.5 kg
(5th-95th percentile 2.6-98.1 kg). 52% were male; 70% White, 28% African
American and 2% unknown race. Indications (Table S1) were mainly
gastroesophageal reflux (18), gastroparesis (9) and headache (8). Twenty
patients received an intravenous bolus (median 0.1 mg/kg), 13 received oral
doses (median 0.1 mg/kg), and 15 received intravenous, oral and
nasogastric/nasojejunal/gastrostomy doses.

The same information is available programmatically via
`readModelDb("Ge_2020_metoclopramide")()$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Ge_2020_metoclopramide.R`. The table below
collects them in one place.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` | log(0.4) 1/h | Table 2 (Ka); Eq. 5 |
| `lcl` | log(19.6) L/h per 70 kg | Table 2 (CL); Eq. 6 |
| `lvc` | log(42.9) L per 70 kg | Table 2 (Vc); Eq. 7 |
| `lq` | log(57.1) L/h per 70 kg | Table 2 (Q); Eq. 8 |
| `lvp` | log(83.9) L per 70 kg | Table 2 (Vp); Eq. 9 |
| `lfdepot` | log(0.97) | Table 2 (F); Eq. 10 |
| `e_wt_cl_q` | 0.75 (fixed) | Methods "Covariate selection"; Results; Eqs. 6 and 8 |
| `e_wt_vc_vp` | 1 (fixed) | Methods "Covariate selection"; Results; Eqs. 7 and 9 |
| `etalcl` | 0.16501 = log(1 + 0.424^2) | Table 2 (IIV CL 42.4 %CV); Eq. 1 |
| `propSd` | 0.333 | Table 2 (proportional error 33.3%) |
| Two-compartment, first-order absorption, no lag | n/a | Results "PopPK model development and evaluation" |
| One Ka and one F for all enteral routes | n/a | Results (route-specific Ka / F did not improve the fit) |
| IV doses into `central`, enteral doses into `depot` | n/a | Methods "PopPK analysis" (routes fitted simultaneously) |
| `Cc = 1000 * central / vc` (ng/mL) | n/a | Methods "Analytical methods" (assay in ng/mL) |

# Deterministic checks

## Steady-state volume

The Discussion reports a typical steady-state volume
Vss = Vc + Vp = 1.81 L/kg. With both volumes linear in weight this is
weight-independent.

```{r vss}
mod <- readModelDb("Ge_2020_metoclopramide")
ini_df <- rxode2::rxode(mod)$iniDf
theta <- setNames(ini_df$est, ini_df$name)
vss_per_kg <- (exp(theta[["lvc"]]) + exp(theta[["lvp"]])) / 70
vss_per_kg
stopifnot(abs(vss_per_kg - 1.81) < 0.005)
```

## Steady-state AUC over one dosing interval (Eq. 4)

The paper computes AUCss,0-6h as Dose / (CL/F) (Eq. 4). For the typical subject
at several body weights, the steady-state profile solved from the ODEs and
integrated by PKNCA must equal F * Dose / CL.

```{r auc-identity}
mod_typ <- rxode2::zeroRe(mod)
wts <- c(3.5, 10, 30, 70)
grid <- seq(0, 6, by = 0.05)
ev_typ <- dplyr::bind_rows(lapply(seq_along(wts), function(i) {
  dplyr::bind_rows(
    data.frame(id = i, time = 0, amt = 0.1 * wts[i], evid = 1L, cmt = "depot",
               ss = 1L, ii = 6),
    data.frame(id = i, time = grid, amt = 0, evid = 0L, cmt = "central",
               ss = 0L, ii = 0)
  ) |>
    dplyr::mutate(WT = wts[i])
}))
sim_typ <- rxode2::rxSolve(mod_typ, events = ev_typ, keep = "WT",
                           rtol = 1e-10, atol = 1e-12,
                           ssRtol = 1e-10, ssAtol = 1e-12, maxsteps = 1e6) |>
  as.data.frame()

conc_typ <- sim_typ |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(Cc = pmax(Cc, 0), treatment = paste0(WT, " kg")) |>
  dplyr::select(id, time, Cc, treatment)
dose_typ <- ev_typ |>
  dplyr::filter(evid == 1) |>
  dplyr::mutate(treatment = paste0(WT, " kg")) |>
  dplyr::select(id, time, amt, treatment)
nca_typ <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_typ, Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(dose_typ, amt ~ time | treatment + id),
  intervals = data.frame(start = 0, end = 6, auclast = TRUE, cmax = TRUE)
))

auc_chk <- as.data.frame(nca_typ$result) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(treatment, auclast = PPORRES) |>
  dplyr::mutate(
    WT = as.numeric(sub(" kg", "", treatment)),
    eq4 = 1000 * exp(theta[["lfdepot"]]) * 0.1 * WT /
      (exp(theta[["lcl"]]) * (WT / 70)^0.75),
    pct_diff = 100 * (auclast / eq4 - 1)
  )
stopifnot(nrow(auc_chk) == length(wts))
auc_chk |>
  dplyr::select(WT, auclast, eq4, pct_diff) |>
  dplyr::rename(
    "Body weight (kg)" = WT,
    "PKNCA AUCss,0-6h (ng*h/mL)" = auclast,
    "Eq. 4 F x Dose / CL (ng*h/mL)" = eq4,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = 3,
               caption = "Typical-value AUCss,0-6h after 0.1 mg/kg q6h.")
# Linear trapezoid on a 0.05 h grid; the residual is integration error only.
stopifnot(max(abs(auc_chk$pct_diff)) < 0.5)
```

# Virtual cohort

Ge 2020 generated 500 virtual patients per postnatal-age group from the PK-Sim
European population. PK-Sim is not available here, so weights are drawn from an
approximate growth curve: the sex-averaged median weight-for-age of the WHO
(0-2 years) and CDC (2-18 years) growth charts, interpolated on age, with
log-normal scatter (SD 0.15 on the log scale). Age is drawn uniformly within
each group. Following the vignette cohort cap, 200 virtual patients are
simulated per age group and regimen.

```{r cohort}
set.seed(2020)
rxode2::rxSetSeed(2020)

# Sex-averaged median weight-for-age (kg): WHO 0-24 months, CDC 2-18 years.
growth <- data.frame(
  age_y = c(0, 1 / 12, 3 / 12, 6 / 12, 1, 2, 3, 5, 8, 10, 12, 14, 16, 18),
  wt_kg = c(3.3, 4.3, 6.1, 7.6, 9.3, 12.0, 14.2, 18.2, 25.6, 32.3, 40.5, 50.0, 58.5, 63.5)
)

age_groups <- data.frame(
  group = c("< 1 month", "1 month to < 2 years", "2 to < 12 years", "12 to < 18 years"),
  lo = c(0, 1 / 12, 2, 12),
  hi = c(1 / 12, 2, 12, 18)
)
age_groups$group <- factor(age_groups$group, levels = age_groups$group)

n_per_group <- 200L
obs_times <- c(0, 0.25, 0.5, seq(1, 6, by = 0.25))

make_cohort <- function(dose_mgkg, id_offset) {
  subj <- dplyr::bind_rows(lapply(seq_len(nrow(age_groups)), function(g) {
    age <- stats::runif(n_per_group, age_groups$lo[g], age_groups$hi[g])
    data.frame(
      group = age_groups$group[g],
      AGE = age,
      WT = stats::approx(growth$age_y, growth$wt_kg, xout = age)$y *
        exp(stats::rnorm(n_per_group, 0, 0.15))
    )
  }))
  subj$id <- id_offset + seq_len(nrow(subj))
  subj$regimen <- paste0(dose_mgkg, " mg/kg q6h")
  doses <- subj |>
    dplyr::mutate(time = 0, amt = dose_mgkg * WT, evid = 1L, cmt = "depot",
                  ss = 1L, ii = 6)
  obs <- subj |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(amt = 0, evid = 0L, cmt = "central", ss = 0L, ii = 0)
  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  make_cohort(0.10, id_offset = 0L),
  make_cohort(0.15, id_offset = 10000L)
)
events$group <- as.character(events$group)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(mod, events = events,
                       keep = c("group", "regimen", "WT"),
                       maxsteps = 1e6) |>
  as.data.frame() |>
  dplyr::mutate(group = factor(group, levels = levels(age_groups$group)))
stopifnot(all(sim$Cc >= -1e-6 * max(sim$Cc, na.rm = TRUE), na.rm = TRUE))
```

The paper reports C~ss,max~ from simulated concentrations at 0.5, 1, 2, 4 and
6 h after the dose (the sampling times of Kearns 1988), and AUC~ss,0-6h~ from
Eq. 4 (F * Dose / CL, with the individual CL).

```{r exposure}
f_oral <- exp(theta[["lfdepot"]])
exposure <- sim |>
  dplyr::group_by(id, group, regimen, WT) |>
  dplyr::summarise(
    cssmax = max(Cc[time %in% c(0.5, 1, 2, 4, 6)]),
    cl = cl[1],
    .groups = "drop"
  ) |>
  dplyr::mutate(
    dose = ifelse(regimen == "0.1 mg/kg q6h", 0.1, 0.15) * WT,
    aucss_eq4 = 1000 * f_oral * dose / cl
  )
stopifnot(nrow(exposure) == 2L * nrow(age_groups) * n_per_group)
```

# Replicate published figures

```{r figure-3, fig.width = 7, fig.height = 4}
ggplot(exposure, aes(WT, cssmax, colour = group)) +
  geom_point(alpha = 0.5, size = 0.8) +
  geom_hline(yintercept = c(26, 94), linetype = "dashed") +
  geom_hline(yintercept = 143) +
  facet_wrap(~regimen) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Body weight (kg)", y = "Css,max (ng/mL)", colour = "Age group",
       title = "Simulated Css,max vs body weight",
       caption = "Replicates Figure 3 of Ge 2020. Dashed: 26-94 ng/mL; solid: 143 ng/mL.")
```

```{r figure-4, fig.width = 7, fig.height = 4}
ggplot(exposure, aes(group, aucss_eq4)) +
  geom_boxplot(outlier.size = 0.6) +
  geom_hline(yintercept = c(115, 374), linetype = "dashed") +
  facet_wrap(~regimen) +
  scale_y_log10() +
  labs(x = NULL, y = "AUCss,0-6h (ng*h/mL)",
       title = "Simulated AUCss,0-6h by age group",
       caption = "Replicates Figure 4 of Ge 2020. Dashed: 115-374 ng*h/mL.") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

# PKNCA validation

PKNCA integrates each simulated steady-state profile (0.25 h grid) over the
0-6 h dosing interval. Its AUC must agree with the paper's Eq. 4 for every
virtual patient, since at steady state the interval AUC equals F * Dose / CL.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(Cc = pmax(Cc, 0),
                treatment = paste(regimen, group, sep = " | ")) |>
  dplyr::select(id, time, Cc, treatment)
dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::mutate(treatment = paste(regimen, group, sep = " | ")) |>
  dplyr::select(id, time, amt, treatment)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id),
  intervals = data.frame(start = 0, end = 6, cmax = TRUE, tmax = TRUE,
                         auclast = TRUE)
))

nca_wide <- as.data.frame(nca_res$result) |>
  dplyr::select(id, treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

chk <- nca_wide |>
  dplyr::inner_join(exposure |> dplyr::select(id, aucss_eq4), by = "id") |>
  dplyr::mutate(pct_diff = 100 * (auclast / aucss_eq4 - 1))
stopifnot(nrow(chk) == nrow(exposure))
# The two sides share each subject's drawn parameters; the difference is
# trapezoid error on the 0.25 h grid (largest for fast-absorbing small infants).
stopifnot(
  abs(median(chk$pct_diff)) < 1,
  max(abs(chk$pct_diff)) < 3
)

nca_wide |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    cmax = median(cmax), tmax = median(tmax), auclast = median(auclast),
    .groups = "drop"
  ) |>
  dplyr::rename(
    "Regimen | age group" = treatment,
    "Median Cmax,ss (ng/mL)" = cmax,
    "Median Tmax,ss (h)" = tmax,
    "Median AUCss,0-6h (ng*h/mL)" = auclast
  ) |>
  knitr::kable(digits = 1, caption = "PKNCA steady-state summary by regimen and age group.")
```

## Comparison against published results

The paper reports no NCA table; its simulation results are the percentage of
virtual patients whose C~ss,max~ lies within 26-94 ng/mL and whose
AUC~ss,0-6h~ lies within 115-374 ng*h/mL (Results "Dose-response simulation
results").

```{r compare}
pct_in <- function(x, lo, hi) 100 * mean(x >= lo & x <= hi)
summ <- function(d) {
  c(cmax = pct_in(d$cssmax, 26, 94), auc = pct_in(d$aucss_eq4, 115, 374))
}
e01 <- dplyr::filter(exposure, regimen == "0.1 mg/kg q6h")
e015 <- dplyr::filter(exposure, regimen == "0.15 mg/kg q6h")
s_all <- summ(e01)
s_inf <- summ(dplyr::filter(e01, group == "1 month to < 2 years"))
s_neo <- summ(dplyr::filter(e015, group == "< 1 month"))
pct_tox_adol <- 100 * mean(
  dplyr::filter(e01, group == "12 to < 18 years")$cssmax > 143
)

comparison <- data.frame(
  quantity = c(
    "0.1 mg/kg, all ages: Css,max in 26-94 ng/mL (%)",
    "0.1 mg/kg, all ages: AUCss,0-6h in 115-374 ng*h/mL (%)",
    "0.1 mg/kg, 1 month to < 2 years: Css,max in range (%)",
    "0.1 mg/kg, 1 month to < 2 years: AUCss,0-6h in range (%)",
    "0.15 mg/kg, < 1 month: Css,max in range (%)",
    "0.15 mg/kg, < 1 month: AUCss,0-6h in range (%)",
    "0.1 mg/kg, 12 to < 18 years: Css,max > 143 ng/mL (%)"
  ),
  published = c(84.3, 75.5, 87.8, 82.6, 94.2, 81.8, 1),
  simulated = c(s_all[["cmax"]], s_all[["auc"]], s_inf[["cmax"]],
                s_inf[["auc"]], s_neo[["cmax"]], s_neo[["auc"]], pct_tox_adol)
)
comparison |>
  dplyr::rename(
    "Quantity" = quantity,
    "Ge 2020" = published,
    "This simulation" = simulated
  ) |>
  knitr::kable(digits = 1, caption = "Percentage of virtual patients within the exposure ranges.")
```

The percentages depend on the virtual cohort's weight distribution as well as
on the model, and the cohort here is an approximation of the PK-Sim population
used by the authors, so the age-group cells (200 virtual patients each) are shown rather than
asserted; the two all-ages percentages are asserted to within 10 percentage
points. A mis-transcribed clearance, bioavailability or unit would move
every exposure by tens of percent; the check below asserts that the median
steady-state exposure of the 0.1 mg/kg regimen sits inside both published
ranges in every age group, as the paper's Figures 3a and 4a show.

```{r compare-gate}
med_by_group <- e01 |>
  dplyr::group_by(group) |>
  dplyr::summarise(cmax = median(cssmax), auc = median(aucss_eq4),
                   .groups = "drop")
med_by_group
# Realised medians (identical at 2 and 16 solver threads): Css,max 32.5 / 40.1 / 53.7 / 60.3
# ng/mL and AUCss 163 / 204 / 279 / 314 ng*h/mL from youngest to oldest. The
# closest margins (neonatal Css,max vs 26, adolescent AUC vs 374) are about 5-6
# standard errors of a 200-subject median at 42% CV.
stopifnot(
  nrow(med_by_group) == 4L,
  all(med_by_group$cmax > 26 & med_by_group$cmax < 94),
  all(med_by_group$auc > 115 & med_by_group$auc < 374)
)

# The two all-ages percentages pool 800 virtual patients (binomial SE about
# 1.3-1.5 percentage points). Realised 85.6% and 77.0% against the published
# 84.3% and 75.5%; a 10-point band is ~7 SE wide. Mutation check on this
# cohort: clearance x 0.75 drops the AUC percentage to 62.6% (fails here), and
# clearance x 1.33 drops the neonatal median Css,max to 24.4 ng/mL (fails the
# median gate above).
stopifnot(
  abs(s_all[["cmax"]] - 84.3) < 10,
  abs(s_all[["auc"]] - 75.5) < 10
)
```

# Assumptions and deviations

- **Virtual cohort.** The authors drew 500 virtual patients per age group from
  the PK-Sim (version 7.0) European population. This vignette uses 200 per
  group with weights from an interpolated sex-averaged WHO/CDC median
  weight-for-age curve and log-normal scatter (SD 0.15). The percentages in the
  comparison table therefore differ somewhat from the published ones; the model
  parameters were not adjusted to reduce the difference.
- **Age groups.** The paper names the groups plotted in Figures 3 and 4 only as
  "term infants to adolescents" matched to the FDA pediatric age groups. The
  groups here (< 1 month, 1 month to < 2 years, 2 to < 12 years, 12 to < 18
  years) follow the FDA definitions and the ranges quoted in the Results.
- **Enteral routes.** The final model uses one Ka and one F for oral,
  nasogastric, nasojejunal and gastrostomy dosing (route-specific values did
  not improve the fit), so all enteral doses go into `depot`. Intravenous
  bolus doses go into `central`; the two continuous infusions in the data set
  are supported by a rate on a `central` dose record.
- **IIV.** Only clearance carries interindividual variability. The reported
  42.4 %CV was converted to a log-normal variance with omega^2 = log(CV^2 + 1)
  = 0.16501.
- **Screened covariates.** Serum creatinine was significant after backward
  elimination but was excluded by the authors (model instability, failed
  covariance step, and loss of significance after removing one subject), and a
  < 1 month age-group effect (about 30% lower CL) was dropped in backward
  elimination. Neither is in the model; both are recorded under
  `covariatesDataExcluded`. Table 1 prints serum creatinine in "mg/mL", which
  is read as mg/dL.
- **Errata.** No correction notice for this article was found in Europe PMC
  as of 2026-09-27.
