---
title: "Gentamicin (Germovsek 2016)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gentamicin (Germovsek 2016)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r meta}
mod_meta <- nlmixr2est::nlmixr(readModelDb("Germovsek_2016_gentamicin"))$meta
```

* Citation: `r mod_meta$reference`
* Description: `r mod_meta$description`
* Article (PubMed): <https://pubmed.ncbi.nlm.nih.gov/27270281/>
* Article (DOI): <https://doi.org/10.1128/AAC.00577-16>
* DDMORE Foundation Model Repository: <https://repository.ddmore.eu/model/DDMODEL00000238>

This vignette validates the packaged `Germovsek_2016_gentamicin` model against
the DDMORE Foundation Model Repository entry **DDMODEL00000238**, the source
from which it was extracted. The Germovsek 2016 publication PDF is not
available on this machine, so the validation strategy follows the F.2
self-consistency recipe from the `extract-literature-model` skill: re-simulate
the bundle's shipped event table and confirm the typical-value trajectory
matches the bundle's NONMEM listing and shipped concentrations.

# Population

The Germovsek 2016 meta-analysis pooled 1,325 gentamicin concentrations from
205 patients (neonates and infants) across three studies: Glasgow (Thomson
1988), Uppsala (Nielsen 2009), and Estonia (unpublished). The full publication
PDF is not on disk, so detailed demographics (age range, weight range, sex
balance, race / ethnicity, indication) could not be cross-checked. The
DDMORE bundle's `Simulated_simdataDDM.csv` uses median covariates from the
original analysis (WT 2.12 kg, PMA 33 weeks, PNA 5.4 days, GA 34 weeks,
CREAT 78 umol/L) -- representative of a preterm neonate around the first
postnatal week.

```{r population}
str(mod_meta$population)
```

# Source trace

Every parameter in the model file's `ini()` block carries an in-file
provenance comment pointing back to the DDMORE bundle. The table below
collects them in one place; THETA / OMEGA / SIGMA values come from the
`FINAL PARAMETER ESTIMATE` block of `Output_real_run35b.lst` after
`MINIMIZATION SUCCESSFUL`, equation forms come from
`Executable_run35b_ddm2.mod` `$PK` / `$DES` / `$ERROR`.

| Parameter / equation                              | Value      | Source location |
|---------------------------------------------------|------------|-----------------|
| `lcl` (TVCL)                                      | log(6.21)  | `.lst` THETA TH 1 (typical CL, L/h, at WT = 70 kg, fully mature) |
| `lvc` (TVV1)                                      | log(26.5)  | `.lst` THETA TH 2 (typical V1, L) |
| `lq`  (TVQ)                                       | log(2.15)  | `.lst` THETA TH 3 (typical Q, L/h) |
| `lvp` (TVV2)                                      | log(21.2)  | `.lst` THETA TH 4 (typical V2, L) |
| `lq2` (TVQ2)                                      | log(0.271) | `.lst` THETA TH 5 (typical Q2, L/h) |
| `lvp2` (TVV3)                                     | log(148)   | `.lst` THETA TH 6 (typical V3, L) |
| `t50_pma` (FIXED)                                 | 55.4 wk    | `.lst` THETA TH 7 (PMA maturation T50 in weeks; FIXED in `.mod`) |
| `hill_pma` (FIXED)                                | 3.33       | `.lst` THETA TH 8 (PMA maturation Hill exponent; FIXED in `.mod`) |
| `e_creat_cl`                                      | -0.130     | `.lst` THETA TH 9 (creatinine power exponent on CL) |
| `p50_pna`                                         | 1.70 d     | `.lst` THETA TH 10 (PNA half-saturation in days) |
| `e_wt_cl` (FIXED)                                 | 0.632      | `.mod` `$PK` `TVCL = THETA(1)*MF*(WTKG/70)**(0.632)` |
| `e_wt_q`  (FIXED)                                 | 0.75       | `.mod` `$PK` `TVQ  = THETA(3)*(WTKG/70)**(0.75)` and `TVQ2` |
| `e_wt_vc` (FIXED)                                 | 1          | `.mod` `$PK` `TVV1 / TVV2 / TVV3 = THETA(*)*(WTKG/70)` |
| `etalcl + etalvc ~ c(0.175, 0.116, 0.112)`        | block(2)   | `.lst` OMEGA(1,1)=0.175; OMEGA(2,1)=0.116; OMEGA(2,2)=0.112 |
| `etalvp ~ 0.132`                                  | 0.132      | `.lst` OMEGA(4,4) (V2) |
| `etalvp2 ~ 0.177`                                 | 0.177      | `.lst` OMEGA(6,6) (V3) |
| `propSd <- sqrt(0.0360)`                          | 0.190      | `.lst` SIGMA(1,1) = 0.0360 (variance) |
| `addSd  <- sqrt(0.0164)`                          | 0.128      | `.lst` SIGMA(2,2) = 0.0164 (variance) |
| `mf <- pma_w^h / (pma_w^h + t50^h)`               | n/a        | `.mod` `$PK` `MF = PMA**HILL/(PMA**HILL+T50**HILL)` |
| `pnaf <- pna_d / (p50 + pna_d)`                   | n/a        | `.mod` `$DES` `PNAF = TCOV1/(P50+TCOV1)` |
| `tcrea <- -2.8488 * pma_w + 166.48`               | n/a        | `.mod` `;TCREA = typical (for PMA) SCr` (Cuzzolin 2006, Rudd 1983) |
| `of <- (CREAT / tcrea)^e_creat_cl`                | n/a        | `.mod` `$DES` `OF = (TCOV2/TCREA)**CRPWR` |
| `cl <- exp(lcl + etalcl) * mf * (WT/70)^0.632 * pnaf * of` | n/a | `.mod` `$PK` `CL = TVCL*EXP(ETA(1)+BOVC)`; PNAF / OF folded in via `K*PNAF*OF` in `$DES` |
| `d/dt(central) ... d/dt(peripheral1) ... d/dt(peripheral2)` | n/a | `.mod` `$DES` `DADT(1) / DADT(2) / DADT(3)` (3-compartment IV) |
| `Cc ~ add(addSd) + prop(propSd)`                  | n/a        | `.mod` `$ERROR` `Y = IPRED*(1+EPS(1)) + EPS(2)` |

# Virtual cohort

The DDMORE bundle ships a 205-subject simulated event table at
`Simulated_simdataDDM.csv`, with all subjects given the same median
covariates from the original analysis. The vignette uses a small
representative cohort (one subject at the bundle's median covariate values
and a few sensitivity variants on PMA / PNA / WT) so the simulation runs
under the pkgdown five-minute budget while still exercising every covariate
effect of interest.

Canonical units are used on input: `WT` in kg (the source data column was in
grams), `PAGE` in months (the source used PMA in weeks), `PNA` in months
(the source used PNA in days), `CREAT` in umol/L. The model converts back
to the source-paper units (weeks, days) at use site so the numerical T50
and P50 values match the `.lst` final estimates.

```{r cohort}
set.seed(20260506)

# Bundle median covariates
ref_cov <- list(
  WT_kg     = 2.12,         # source dataset uses 2120 g
  PMA_weeks = 33,           # canonical PAGE = 33 / 4.345 months
  PNA_days  = 5.4,          # canonical PNA  = 5.4 / 30.4375 months
  CREAT     = 78            # umol/L
)

cohorts <- tibble::tibble(
  cohort       = c("median (simdata)", "term newborn", "older infant"),
  WT_kg        = c(2.12, 3.50, 6.00),
  PMA_weeks    = c(33,   40,   60),
  PNA_days     = c(5.4,  3,    150),
  CREAT_uM     = c(78,   55,   30)
)

# Dosing -- gentamicin 5 mg/kg IV q24h x 5 doses, 5-minute infusions.
# `rate` chosen so amt / rate = 1/12 h = 5 minutes.
dose_times    <- c(0, 24, 48, 72, 96)
sample_times  <- c(0.083, 0.5, 1, 2, 4, 8, 12, 24,
                   24.083, 24.5, 25, 26, 28, 32, 36, 48,
                   72.5, 96.5, 100, 110, 120)

make_subject <- function(idx, row) {
  amt   <- 5 * row$WT_kg                   # mg
  rate  <- amt * 12                        # mg/h => 1/12 h infusion
  doses <- tibble::tibble(
    id   = idx, time = dose_times,
    evid = 1L,  amt = amt, rate = rate, dv = NA_real_
  )
  obs <- tibble::tibble(
    id   = idx, time = sample_times,
    evid = 0L,  amt = NA_real_, rate = NA_real_, dv = NA_real_
  )
  bind_rows(doses, obs) |>
    mutate(
      cohort = row$cohort,
      WT     = row$WT_kg,
      PAGE   = row$PMA_weeks / 4.345,
      PNA    = row$PNA_days  / 30.4375,
      CREAT  = row$CREAT_uM
    ) |>
    arrange(time, desc(evid))
}

events <- bind_rows(lapply(seq_len(nrow(cohorts)), function(i) {
  make_subject(idx = i, row = cohorts[i, ])
}))

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

```{r simulate-typical}
mod          <- readModelDb("Germovsek_2016_gentamicin")
mod_typical  <- rxode2::zeroRe(mod)

sim_typical <- rxode2::rxSolve(
  object = mod_typical, events = events,
  keep   = c("cohort")
) |>
  as.data.frame()
```

```{r simulate-stochastic}
# Replicate each cohort 30 times to build a stochastic VPC sample.
n_replicates <- 30L
events_stoch <- bind_rows(lapply(seq_len(n_replicates), function(rep_idx) {
  events |>
    mutate(id = id + (rep_idx - 1L) * nrow(cohorts))
}))
stopifnot(!anyDuplicated(unique(events_stoch[, c("id", "time", "evid")])))

sim_stoch <- rxode2::rxSolve(
  object = mod, events = events_stoch,
  keep   = c("cohort")
) |>
  as.data.frame()
```

# F.2 self-consistency check against the DDMORE bundle

The bundle's `Simulated_simdataDDM.csv` records the simulated `DV` column
under the model's typical population at the median covariates above. A
representative trace from the file (subject 1001) had `DV ~= 7.02 mg/L`
30 min after a 6 mg dose at 12 h, and `DV ~= 2.21 mg/L` at 58.5 h after
4 x maintenance doses; the values include residual error.
The block below confirms the packaged nlmixr2 model reproduces those
DDMORE-bundle predictions for the same subject and dose history.

```{r f2-resimulate}
spot_events <- tibble::tibble(
  id   = 1L,
  time = c(0, 11.5, 12, 12.5, 24, 36, 48, 58.5),
  evid = c(1, 0, 1, 0, 1, 1, 1, 0),
  amt  = c(6, 0, 6, 0, 4, 4, 4, 0),
  rate = c(72, 0, 72, 0, 48, 48, 48, 0),
  WT   = 2.12,
  PAGE = 33 / 4.345,
  PNA  = 5.4 / 30.4375,
  CREAT = 78
)

spot_typical <- rxode2::rxSolve(
  object = mod_typical, events = spot_events
) |>
  as.data.frame() |>
  filter(time %in% c(11.5, 12.5, 58.5)) |>
  select(time, Cc_typical = Cc)

bundle_dv <- tibble::tibble(
  time         = c(11.5, 12.5, 58.5),
  Cc_simdataDV = c(1.109, 7.018, 2.214)   # from Simulated_simdataDDM.csv id 1001
)

knitr::kable(
  spot_typical |> left_join(bundle_dv, by = "time") |>
    mutate(rel_diff_pct = round(100 * (Cc_typical - Cc_simdataDV) / Cc_simdataDV, 1)),
  digits = 3,
  caption = "Typical-value re-simulation vs DDMORE bundle DV (residual error included in bundle DV)."
)
```

The typical-value Cc and the residual-error-laden DDMORE `DV` agree to within
the magnitude of the residual error reported in the model
(`addSd ~= 0.13 mg/L` and `propSd ~= 19%`), which is the expected outcome of
F.2 self-consistency.

# Trajectories across the virtual cohort

```{r trajectories}
sim_typical |>
  filter(time > 0) |>
  ggplot(aes(time, Cc, colour = cohort, group = cohort)) +
  geom_line() +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Gentamicin concentration (mg/L)",
    title = "Typical-value trajectories across the virtual cohort",
    subtitle = "Germovsek 2016 / DDMODEL00000238 -- 5 mg/kg IV q24h x 5 doses",
    colour = NULL
  ) +
  theme_minimal()
```

```{r vpc}
sim_stoch |>
  filter(time > 0) |>
  group_by(cohort, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.20) +
  geom_line() +
  facet_wrap(~ cohort) +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Gentamicin concentration (mg/L)",
    title = "Simulated VPC by cohort",
    subtitle = "Median (line) and 5-95% interval (ribbon); IIV + residual error included"
  ) +
  theme_minimal()
```

# PKNCA on the simulated cohort

PKNCA is run on the typical-value cohort over the first dosing interval
(0-24 h) and the steady-state interval (96-120 h). The Germovsek 2016
publication is not on disk, so the simulated NCA values cannot be compared
side-by-side against published Cmax / AUC tables; they are reported here as
a sanity check that the simulation pipeline produces NCA values in the
expected clinical range for gentamicin (Cmax in the 5-15 mg/L range,
trough in the < 2 mg/L range for once-daily neonatal dosing).

```{r pknca}
sim_for_nca <- sim_typical |>
  filter(!is.na(Cc), Cc > 0) |>
  select(id, time, Cc, cohort) |>
  as.data.frame()

doses_for_nca <- events |>
  filter(evid == 1L) |>
  select(id, time, amt, cohort) |>
  as.data.frame()

conc_obj <- PKNCA::PKNCAconc(
  data    = sim_for_nca,
  formula = Cc ~ time | cohort + id,
  concu   = "mg/L",
  timeu   = "hr"
)
dose_obj <- PKNCA::PKNCAdose(
  data    = doses_for_nca,
  formula = amt ~ time | cohort + id,
  doseu   = "mg"
)

intervals <- data.frame(
  start    = c(0, 96),
  end      = c(24, 120),
  cmax     = TRUE,
  tmax     = TRUE,
  auclast  = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- suppressWarnings(PKNCA::pk.nca(nca_data))

knitr::kable(
  summary(nca_res),
  caption = "Simulated NCA parameters by cohort (PKNCA)."
)
```

## Comparison against published NCA

The Germovsek 2016 publication PDF is not on disk under
`/home/bill/github/mab_human_consensus/literature/`, so the in-paper NCA
tables (Cmax, AUC, trough by PMA / PNA stratum) cannot be reproduced here.
This is the F.2 substitute path described in the
`extract-literature-model` skill (`references/ddmore-source.md` Section 
"Validation strategy by model type"). Operator follow-up: pull the
publication PDF and compare PKNCA outputs above against any in-paper
Cmax / AUC / Cmin values reported by Germovsek et al.

# Assumptions and deviations

- **Parameter values come from the bundle's `Output_real_run35b.lst`
  `FINAL PARAMETER ESTIMATE` block.** The `.mod` `$THETA / $OMEGA / $SIGMA`
  blocks are initial estimates and are NOT used for parameter values.
  T50 and Hill on PMA were FIXED in the source ($THETA `55.4 FIX`, `3.33 FIX`)
  and are wrapped in `fixed(...)` in `ini()`.

- **`MINIMIZATION SUCCESSFUL HOWEVER, PROBLEMS OCCURRED WITH THE MINIMIZATION.`**
  The `.lst` flagged minimization difficulties: "REGARD THE RESULTS OF THE
  ESTIMATION STEP CAREFULLY, AND ACCEPT THEM ONLY AFTER CHECKING THAT THE
  COVARIANCE STEP PRODUCES REASONABLE OUTPUT." The DDMORE bundle's RDF
  metadata still asserts `model-implementation-conforms-to-literature-controlled = "Yes"`
  with no source discrepancies; the values are extracted as published, but
  reviewers using this model for re-fitting should be aware of the source
  message and cross-check the covariance step on the original data.

- **Inter-occasion variability (IOV) was dropped.** The source `.mod`
  defined a per-occasion CL random effect via 22 ETAs in BLOCK(1) SAME
  blocks (`OCC = 1..22`), each with variance 0.0141 (~= 12% CV). The
  packaged nlmixr2 model omits IOV because (a) it is small relative to
  the inter-individual CL variance (0.175, ~= 42% CV), (b) nlmixr2's
  occasion-effect ergonomics differ from NONMEM's `BLOCK(1) SAME` idiom,
  and (c) the model is intended for typical-value and population
  simulation, not per-occasion re-estimation. Users fitting on real data
  should reintroduce occasion-level variability through their fitting
  workflow.

- **Time-varying covariate compartments dropped.** The source `.mod`
  carried PNA and CREAT through dummy compartments (`A_0(4) = PNA`,
  `DADT(4) = SL1`, `TCOV1 = A(4)` and similarly for CREAT) so NONMEM
  could linearly interpolate covariate values between observation events.
  The packaged nlmixr2 model uses rxode2's native time-varying-covariate
  handling: `PAGE`, `PNA`, and `CREAT` are read directly from the event
  table at each integration step (last-observation-carried-forward by
  default; rxode2 also supports `linear` interpolation via the model
  options if a user prefers continuous interpolation). The two approaches
  are mathematically equivalent at observation time points; mid-interval
  values differ only when the covariate column itself changes between
  events.

- **Canonical PAGE / PNA units (months) vs source-paper units
  (weeks / days).** The covariate-columns register fixes PAGE in months
  and PNA in months. The Germovsek 2016 source uses PMA in weeks (with
  T50 = 55.4 weeks, Hill = 3.33) and PNA in days (with P50 = 1.70 days).
  To preserve the `.lst` final-estimate values verbatim, the model file
  reports T50 and P50 on the source-paper scale and converts the
  canonical-input PAGE / PNA back to weeks / days at use sites
  (`pma_weeks <- PAGE * 4.345`, `pna_days <- PNA * 30.4375`). The Hill
  function and the PNAF saturation are dimensionally homogeneous, so this
  internal rescaling is exact; users supply PAGE in months and PNA in
  months as documented in the canonical register.

- **Body weight unit conversion.** The source dataset stores `WT` in
  grams (g) and divides by 1000 inside `$PK` (`WTKG = WT/1000`). The
  packaged nlmixr2 model expects `WT` in kg (the canonical convention),
  matching the rest of `inst/modeldb/`.

- **Allometric exponents (1, 0.75, 0.632) are FIXED.** None of the
  exponents (CL, Q/Q2, V1/V2/V3 weight scaling) were estimated as
  `THETA(...)`s -- they are hard-coded in the `.mod` `$PK` block. They
  are wrapped in `fixed(...)` in `ini()` so estimation does not perturb
  them.

- **Sex (`GIRL`), gestational age (`GA`), study indicator (`STUDY`),
  and occasion (`OCC`) are present in the source `$INPUT` but not used
  in `$PK` / `$DES`.** They are not declared as covariates in the
  packaged model.

- **Validation strategy is F.2 self-consistency** (per
  `references/ddmore-source.md` Section  "Validation strategy by model type"
  decision tree, leaf 1: linked publication exists but is not on disk).
  PKNCA values shown above are informational; comparison against the
  Germovsek 2016 published NCA / population-prediction figures could
  not be performed.

- **Missing-CREAT imputation.** The source `.mod` substitutes the typical
  PMA-dependent SCr (`TCREA = -2.8488 * PMA_weeks + 166.48`) when the
  observed CREAT is missing (coded `-99`). The nlmixr2 model assumes the
  user supplies a non-missing CREAT; users with missing serum creatinine
  should impute with the same TCREA formula before passing CREAT to the
  model.
