---
title: "Dalbavancin (Giacometti 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dalbavancin (Giacometti 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Article: [Entropy 2025;27(6):602](https://doi.org/10.3390/e27060602)
* Analysis code (weights not distributed): <https://github.com/TommyGiak/pharmacoNODE>

Two model files come from this paper, both fitted to the same 218-patient
cohort and differing only in how they were estimated:

```{r models}
mod_nlme <- readModelDb("Giacometti_2025_dalbavancin_nlme")
mod_nlls <- readModelDb("Giacometti_2025_dalbavancin_nlls")

ui_nlme <- rxode2::rxode(mod_nlme)
ui_nlls <- rxode2::rxode(mod_nlls)

ui_nlme$reference
```

* `Giacometti_2025_dalbavancin_nlme` -- the nonlinear mixed-effects (Monolix,
  maximum likelihood) fit.
* `Giacometti_2025_dalbavancin_nlls` -- the naive-pooled nonlinear
  least-squares fit, packaged as the paper's **negative comparator**.

```{r descriptions}
ui_nlme$description
ui_nlls$description
```

# What this paper is, and what is extractable from it

Giacometti 2025 is a three-way methodological comparison. It asks whether a
**Neural ODE** (NODE) -- an ODE whose right-hand side is replaced by a
feed-forward neural network -- can predict dalbavancin concentrations as well
as, or better than, the two conventional approaches. All three arms were
trained and evaluated on one real therapeutic-drug-monitoring cohort under
6-fold cross-validation.

| Arm | Paper's Table 4 column | Extractable? |
|---|---|---|
| Naive-pooled nonlinear least squares, 2-compartment | "Two-Compartment" | Yes -> `Giacometti_2025_dalbavancin_nlls` |
| Nonlinear mixed effects (Monolix), 2-compartment | "NLME Model" | Yes -> `Giacometti_2025_dalbavancin_nlme` |
| Neural ODE (with and without covariates) | none | **No** -- see below |
| Prior literature values | "Literature [15]" | Already in the library as `Cojutti_2024_dalbavancin` |

**The Neural ODE itself is not extractable, and that is a reporting limit
rather than a choice made here.** Its right-hand side is a feed-forward network
with two (no-covariate) or three (covariate) hidden layers of 20 softplus units
over four or nine inputs, plus a separate volume network; that is several
hundred trained weights, and the paper publishes none of them. The only NODE
quantities given numerically are the learned distribution volumes,
`V_NODE = 5.7 +/- 0.1 L` (Equation 5) and `V_NODE,cov = 5.2 +/- 0.8 L`
(Equation 6). The analysis code is public but the trained weights and the
patient data are not, and nlmixr2 / rxode2 has no construct for a layered
network inside `model()` in any case. No weight is invented here; the two
conventional arms are extracted in full and the NODE volume is used below only
as an independent cross-check on `V1`.

# Population

The cohort is 218 adults who received dalbavancin under therapeutic drug
monitoring at the IRCCS Azienda Ospedaliero-Universitaria di Bologna between
April 2021 and December 2024 (Methods 2.1). Dalbavancin was given for skin and
soft tissue infection and, as a second-line agent, for staphylococcal bone and
joint infection, vascular prosthetic joint infection and endocarditis. There
were 703 recorded intravenous administrations spanning 350 to 1500 mg, and 669
plasma concentration measurements. Sampling is explicitly sparse and irregular:
patients frequently received several doses before any concentration was drawn.

Baseline covariates (Table 1) were age 18 to 92 years (mean 64, SD 16), height
145 to 190 cm (mean 171, SD 9), weight 40 to 140 kg (mean 77, SD 16) and serum
creatinine 19 to 411 umol/L (mean 94, SD 48). Sex was 145 male and 73 female
(33.5% female). Race and ethnicity are not reported.

```{r population}
str(ui_nlme$population, max.level = 1)
```

**No covariate is carried by either model, deliberately.** Methods 2.2 states
that "the covariates were not included in the NLME and two-compartment models
because, when tested, their inclusion did not improve the results", and
Appendix B reports the supporting test: the residual distribution of the
Monolix fit with per-fold best covariates was indistinguishable from the
covariate-free fit (Kolmogorov-Smirnov p = 0.45), so the authors kept the
simpler model as "a robust and scalable version" that "does not rely on the
specific choices of the covariates to include for each parameter, which vary a
lot depending on the specific iterations of the cross-validation." All five
screened covariates are recorded in each model's `covariatesDataExcluded`
metadata so the provenance of the screen is preserved.

# Source trace

Every value below is a plus-or-minus MEAN over the six cross-validation folds,
and the quoted spread is the standard deviation **across folds** (Table 4
caption). It is not a standard error, not an RSE, and not a between-subject
variance -- a point that matters for the inter-individual-variability gap
discussed under Assumptions and deviations.

| Equation / parameter | NLME value | NLLS value | Source location |
|---|---|---|---|
| `d/dt(central)`, `d/dt(peripheral1)` | n/a | n/a | Equation 2a; schematic Figure 2; Monolix configuration in Methods 2.2 |
| `Cc <- central / vc` | n/a | n/a | Equation 2a (the state divided by `V1`); units from Table 4 row labels |
| `lcl` (CL, L/h) | 0.0367 (SD 0.0006) | 0.054 (SD 0.002) | Table 4, row "Cl (L/h)" |
| `lvc` (V1, L) | 6.32 (SD 0.15) | 5.3 (SD 0.4) | Table 4, row "V 1 (L)" |
| `lq` (Q, L/h) | 0.028 (SD 0.003) | 0.42 (SD 0.14) | Table 4, row "Q (L/h)" |
| `lvp` (V2, L) | 13.9 (SD 0.6) | 8.8 (SD 1.5) | Table 4, row "V 2 (L)" |
| `propSd` | 0 (fixed) | 0 (fixed) | Not reported anywhere in the article |
| `addSd` | 0 (fixed) | 0 (fixed) | Not reported anywhere in the article |
| inter-individual variability | not declared | structurally absent | Table 4 prints no omega block; Methods 2.2 for the NLLS arm |
| efficacy threshold 8.04 mg/L | n/a | n/a | Methods 2.1 and Figure 3 caption |
| `V_NODE = 5.7 L` (cross-check only) | n/a | n/a | Equation 5 |

## Reading Equation 2a

Equation 2a is printed as

```
dCc/dt = -Cc*Cl/V1 - Cc*Q/V1 + Cp*Q/V2
dCp/dt =  Cc*Q/V1  - Cp*Q/V2
```

with the surrounding prose naming `Cc` and `Cp` "the concentrations of the two
compartments". Read literally on concentrations this system does **not**
conserve mass, because the peripheral compartment's influx and efflux would be
divided by different volumes. Substituting amounts (`Cc -> A1`, `Cp -> A2`)
makes it the canonical mass-conserving two-compartment system, with
`d(A1 + A2)/dt = -(CL/V1) * A1`. The equation is therefore read on **amounts**
and the prose labelling is the slip. Two independent facts confirm this
reading: the NLME arm was fitted in Monolix configured as "two-compartment
distribution, infusion, no delay and linear elimination" (Methods 2.2), which
is the standard mass-conserving form; and Figure 2 describes the central
compartment as "the only compartment that can eliminate the antibiotic at rate
Cl". The mass-balance check below is the mechanical version of this argument.

# Virtual cohort

**These models are fully deterministic.** Neither declares an `eta`, and both
residual-error terms are `fixed(0)` because the paper reports no variance
component of any kind. There is therefore no simulated cohort, no RNG, and no
between-subject spread to summarise -- so the assertions in this vignette are
exact numerical identities rather than tolerance bands on a random draw, and
they are written tight on purpose. (The usual caution about assertions on
cohort-derived quantities being irreproducible across solver thread counts does
not apply to anything below; there is nothing random to draw.)

The "cohort" is instead one subject per dose level, spanning the dose range
the paper reports (350 to 1500 mg, Methods 2.1), simulated under both models.

```{r cohort}
DOSES <- c(350, 500, 1000, 1500)

# The paper states "infusion, no delay" (Methods 2.2) but never gives the
# infusion duration; 30 min is the dalbavancin label duration. See Assumptions.
T_INF <- 0.5

# Dense early, coarse late: the distribution phase is resolved finely so NCA
# recovers Cmax and AUC exactly, and the record runs to 5000 h (about 7.5
# NLME terminal half-lives) so the AUC extrapolation is negligible.
grid <- sort(unique(c(
  seq(0,    2,    by =  0.05),
  seq(2,    48,   by =  0.5),
  seq(48,   720,  by =  4),
  seq(720,  5000, by = 20)
)))

make_arm <- function(dose, id) {
  as.data.frame(
    rxode2::et(amt = dose, dur = T_INF, cmt = "central", id = id) |>
      rxode2::et(grid, cmt = "central", id = id)
  ) |>
    dplyr::mutate(dose_mg = dose)
}

events <- dplyr::bind_rows(
  lapply(seq_along(DOSES), function(i) make_arm(DOSES[i], i))
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
c(rows = nrow(events), subjects = length(unique(events$id)))
```

Observation rows carry `cmt = "central"` -- the ODE state, never the
observable name `Cc`. rxode2 returns `Cc` as an output column at those rows.

# Simulation

```{r simulate}
solve_one <- function(mod, label) {
  s <- rxode2::rxSolve(
    mod, events,
    keep       = c("dose_mg"),
    returnType = "data.frame",
    atol = 1e-14, rtol = 1e-12
  )
  if (is.null(s$id)) s$id <- 1L
  s$model <- label
  s
}

sim <- dplyr::bind_rows(
  solve_one(mod_nlme, "NLME"),
  solve_one(mod_nlls, "NLLS")
) |>
  dplyr::mutate(
    arm = paste0(model, " ", dose_mg, " mg"),
    # Disjoint ids across the two models so PKNCA never merges subjects.
    nca_id = id + ifelse(model == "NLLS", length(DOSES), 0L)
  )

stopifnot(all(sim$Cc >= 0), !anyNA(sim$Cc))
```

# Replicate published figures

## Figure 3 -- the two conventional arms against the efficacy threshold

Figure 3 of Giacometti 2025 overlays the NODE, NLME and two-compartment
predictions for one test patient, with the 8.04 mg/L efficacy threshold drawn
as a red dashed line. The NODE curve cannot be reproduced (no weights), so the
panel below shows the two conventional arms after a single 1500 mg infusion.
The qualitative feature the paper reports is visible directly: the two curves
agree through the first days and then diverge, with the least-squares fit
falling away faster.

```{r figure-3}
sim |>
  dplyr::filter(dose_mg == 1500, time > 0, time <= 1344) |>
  ggplot(aes(time / 24, Cc, colour = model)) +
  geom_hline(yintercept = 8.04, linetype = "dashed", colour = "red") +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(
    x = "Time since dose (days)", y = "Total plasma dalbavancin (mg/L)",
    colour = "Estimation",
    title = "Single 1500 mg infusion, typical-value prediction",
    caption = paste(
      "Structure of Figure 3 of Giacometti 2025 (NODE arm omitted: weights not published).",
      "Red dashed line is the 8.04 mg/L efficacy threshold."
    )
  )
```

## Figure 4 -- the long-time bias of the least-squares fit

Figure 4 plots residuals against time since the most recent dose and reports
"a bias in long time predictions for the two-compartment model"; the Discussion
states that "the two-compartment model systematically underestimates long-term
drug concentrations". Observed data are not available, so the panel below shows
the model-versus-model relative difference, which is the deterministic core of
that claim: the least-squares fit sits above the NLME fit early (it absorbs
drug into the peripheral compartment much faster) and below it in the far tail.

```{r figure-4}
rel <- sim |>
  dplyr::filter(dose_mg == 1500, time > T_INF) |>
  dplyr::select(time, model, Cc) |>
  tidyr::pivot_wider(names_from = model, values_from = Cc) |>
  dplyr::mutate(rel_pct = 100 * (NLLS - NLME) / NLME)

ggplot(rel, aes(time / 24, rel_pct)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_line(linewidth = 0.8) +
  labs(
    x = "Time since dose (days)",
    y = "NLLS relative to NLME (%)",
    title = "Relative difference between the two conventional arms",
    caption = "Deterministic analogue of Figure 4 of Giacometti 2025."
  )
```

# Transcription gate: `ini()` against Table 4

Table 4 is restated here as literals and compared against what the two model
files actually carry. This is the only check in this vignette that can catch a
mistranscribed parameter -- every closed-form check further down reads the same
`ini()` values on both sides of the comparison, so it validates structure, not
transcription.

```{r transcription-gate}
# Giacometti 2025 Table 4, typed from the article. Columns are the paper's
# own: "Two-Compartment" (naive-pooled NLLS) and "NLME Model".
table4 <- tibble::tribble(
  ~param, ~nlls,  ~nlme,
  "lcl",   0.054,  0.0367,   # Cl (L/h)
  "lvc",   5.3,    6.32,     # V 1 (L)
  "lq",    0.42,   0.028,    # Q (L/h)
  "lvp",   8.8,    13.9      # V 2 (L)
)

packaged <- data.frame(
  param = table4$param,
  nlls  = exp(ui_nlls$theta[table4$param]),
  nlme  = exp(ui_nlme$theta[table4$param])
)

# Exact equality up to the double-precision round trip through log()/exp().
stopifnot(
  isTRUE(all.equal(packaged$nlls, table4$nlls, tolerance = 1e-12)),
  isTRUE(all.equal(packaged$nlme, table4$nlme, tolerance = 1e-12))
)

# Both residual terms and every variance component must be absent / zero: the
# article reports none, and inventing one is the failure mode this guards.
# `neta1` is asserted PRESENT first -- if the column ever disappears,
# `all(is.na(NULL))` is `all(logical(0))` is TRUE, i.e. a gate that cannot go
# red.
stopifnot(
  "neta1" %in% names(ui_nlme$iniDf), "neta1" %in% names(ui_nlls$iniDf),
  nrow(ui_nlme$iniDf) > 0L, nrow(ui_nlls$iniDf) > 0L
)
stopifnot(
  ui_nlme$theta[["propSd"]] == 0, ui_nlme$theta[["addSd"]] == 0,
  ui_nlls$theta[["propSd"]] == 0, ui_nlls$theta[["addSd"]] == 0,
  all(is.na(ui_nlme$iniDf$neta1)), all(is.na(ui_nlls$iniDf$neta1))
)

data.frame(
  Parameter        = c("CL (L/h)", "V1 (L)", "Q (L/h)", "V2 (L)"),
  `Table 4 NLLS`   = table4$nlls,  `Packaged NLLS` = packaged$nlls,
  `Table 4 NLME`   = table4$nlme,  `Packaged NLME` = packaged$nlme,
  check.names      = FALSE
) |>
  knitr::kable(caption = "Table 4 of Giacometti 2025 vs the packaged `ini()` values.")
```

# Closed-form verification

Three structural checks. Each goes red on a mis-signed or misrouted ODE term, a
wrong volume in `Cc`, or a wrong micro-constant. All are exact identities, not
tolerance bands.

## 1. The solved ODE against the analytical two-compartment infusion solution

```{r verify-analytic}
micro <- function(ui) {
  th  <- ui$theta
  cl  <- exp(th[["lcl"]]); vc <- exp(th[["lvc"]])
  q   <- exp(th[["lq"]]);  vp <- exp(th[["lvp"]])
  k10 <- cl / vc; k12 <- q / vc; k21 <- q / vp
  s   <- k10 + k12 + k21
  d   <- sqrt(s^2 - 4 * k10 * k21)
  list(cl = cl, vc = vc, q = q, vp = vp, k21 = k21,
       alpha = (s + d) / 2, beta = (s - d) / 2,
       vss = vc + vp, t_half_beta = log(2) / ((s - d) / 2))
}

# Analytical Cc(t) for a single zero-order infusion into a 2-compartment
# system (dose in mg, volumes in L, so Cc is mg/L).
conc_analytic <- function(p, dose, t_inf, t) {
  k0 <- dose / t_inf
  A  <- (k0 / p$vc) * (p$k21 - p$alpha) / ((p$beta  - p$alpha) * p$alpha)
  B  <- (k0 / p$vc) * (p$k21 - p$beta)  / ((p$alpha - p$beta)  * p$beta)
  ifelse(
    t <= t_inf,
    A * (1 - exp(-p$alpha * t)) + B * (1 - exp(-p$beta * t)),
    A * (1 - exp(-p$alpha * t_inf)) * exp(-p$alpha * (t - t_inf)) +
      B * (1 - exp(-p$beta  * t_inf)) * exp(-p$beta  * (t - t_inf))
  )
}

pars <- list(NLME = micro(ui_nlme), NLLS = micro(ui_nlls))

analytic_chk <- sim |>
  dplyr::filter(time > 0) |>
  dplyr::rowwise() |>
  dplyr::mutate(Cc_ref = conc_analytic(pars[[model]], dose_mg, T_INF, time)) |>
  dplyr::ungroup() |>
  dplyr::filter(Cc_ref > 1e-8) |>
  dplyr::mutate(pct = 100 * (Cc - Cc_ref) / Cc_ref)

max_analytic_pct <- max(abs(analytic_chk$pct))

# Realised 1.2e-9 %. The solver is run at atol 1e-14 / rtol 1e-12, so this is
# machine-level agreement and 0.01% leaves five orders of magnitude of headroom.
#
# SCOPE OF THIS GATE. Both sides read CL, V1, Q and V2 from the model's own
# ini(), so it is a STRUCTURAL check: it goes red on a mis-signed or misrouted
# ODE term, on `Cc` divided by the wrong volume, and on a wrong micro-constant.
# It CANNOT go red on a mistranscribed Table 4 value, because a wrong theta
# moves both sides together. The same caveat applies to the mass-balance,
# half-life and PKNCA checks below. Transcription is guarded separately, by the
# executable Table 4 comparison in the next section.
stopifnot(max_analytic_pct < 0.01)
signif(max_analytic_pct, 3)
```

## 2. Mass balance -- `AUC[0,T] * CL == dose in - amount remaining`

This identity holds at **any** T, with no steady-state or terminal-phase
assumption, and it fails if `Cc` is divided by the wrong volume or if either
inter-compartmental term is mis-signed.

```{r verify-mass-balance}
mass_balance <- sim |>
  dplyr::filter(time > 0) |>
  dplyr::group_by(model, dose_mg) |>
  dplyr::arrange(time, .by_group = TRUE) |>
  dplyr::group_modify(function(d, key) {
    cl <- pars[[key$model]]$cl
    purrr_rows <- lapply(c(168, 720, 5000), function(Tend) {
      x   <- d[d$time <= Tend, ]
      auc <- sum(diff(x$time) * (utils::head(x$Cc, -1) + utils::tail(x$Cc, -1)) / 2)
      rem <- utils::tail(x$central, 1) + utils::tail(x$peripheral1, 1)
      data.frame(
        Tend      = Tend,
        eliminated = auc * cl,
        expected   = key$dose_mg - rem
      )
    })
    dplyr::bind_rows(purrr_rows)
  }) |>
  dplyr::ungroup() |>
  dplyr::mutate(pct = 100 * (eliminated - expected) / expected)

max_mb_pct <- max(abs(mass_balance$pct))

# Realised 0.0043%: pure trapezoidal error of the observation grid, not model
# error. 0.05% admits that and still goes red on a structural mistake.
stopifnot(max_mb_pct < 0.05)
signif(max_mb_pct, 3)
```

## 3. Terminal half-life recovered from the simulated tail

```{r verify-half-life}
half_life_chk <- sim |>
  dplyr::filter(dose_mg == 1500, time >= 3000, time <= 5000, Cc > 0) |>
  dplyr::group_by(model) |>
  dplyr::summarise(
    t_half_sim = log(2) / -stats::coef(stats::lm(log(Cc) ~ time))[2],
    .groups = "drop"
  ) |>
  dplyr::mutate(
    t_half_closed = vapply(model, function(m) pars[[m]]$t_half_beta, numeric(1)),
    pct           = 100 * (t_half_sim - t_half_closed) / t_half_closed
  )

# Fitted well past washout of the alpha phase (>= 3000 h is 48 NLLS and 4.5
# NLME alpha half-lives), so this recovers log(2)/beta to machine precision.
stopifnot(max(abs(half_life_chk$pct)) < 0.1)

half_life_chk |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) signif(x, 5))) |>
  dplyr::rename(
    "Estimation"                 = model,
    "t1/2 from simulated tail (h)" = t_half_sim,
    "t1/2 closed form (h)"         = t_half_closed,
    "Difference (%)"               = pct
  ) |>
  knitr::kable(caption = "Terminal half-life: simulated tail slope vs log(2)/beta.")
```

# PKNCA validation

The paper reports no NCA parameters, so the reference column is the closed-form
value each parameter must take for a single intravenous dose:
`AUC(0-inf) = Dose / CL` exactly, and `t1/2 = log(2) / beta`. That makes this a
real gate rather than a restatement of the model.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(nca_id, time, Cc, arm)

# Guarantee a time-zero record per (arm, id). For an intravenous dose the
# pre-dose concentration is 0. Filter above is !is.na() only -- adding
# `time > 0` or `Cc > 0` would drop this anchor and trigger PKNCA's
# "AUC range starting (0) before the first measurement" warning.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(nca_id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(nca_id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(nca_id, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + nca_id)

# Dose records come from `events`, NOT from `sim`: the rxSolve output frame
# carries no `evid` / `amt` columns. The same event table drove both solves, so
# it is replicated per model with the matching disjoint `nca_id` offset.
dose_df <- dplyr::bind_rows(
  events |> dplyr::mutate(model = "NLME", nca_id = id),
  events |> dplyr::mutate(model = "NLLS", nca_id = id + length(DOSES))
) |>
  dplyr::filter(evid != 0, !is.na(amt)) |>
  dplyr::mutate(arm = paste0(model, " ", dose_mg, " mg")) |>
  dplyr::select(nca_id, time, amt, arm) |>
  dplyr::distinct()

stopifnot(nrow(dose_df) == 2L * length(DOSES),
          setequal(dose_df$arm, unique(sim_nca$arm)))

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + nca_id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

nca_res <- suppressWarnings(
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals),
                verbose = FALSE)
)
```

## Comparison against closed-form reference values

```{r nca-compare}
reference <- sim |>
  dplyr::distinct(arm, model, dose_mg) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    cmax       = conc_analytic(pars[[model]], dose_mg, T_INF, T_INF),
    tmax       = T_INF,
    aucinf.obs = dose_mg / pars[[model]]$cl,
    half.life  = pars[[model]]$t_half_beta
  ) |>
  dplyr::ungroup() |>
  dplyr::select(arm, cmax, tmax, aucinf.obs, half.life)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = reference,
  by            = "arm",
  units         = c(cmax = "mg/L", tmax = "h",
                    aucinf.obs = "mg*h/L", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated NCA vs closed-form reference (Dose/CL for AUC(0-inf),",
    "log(2)/beta for t1/2, analytical solution for Cmax).",
    "* marks a >20% difference; none is expected or observed."
  ),
  align = c("l", rep("r", ncol(cmp) - 1))
)
```

```{r nca-gate}
nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(arm, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::left_join(
    reference |> dplyr::rename(cmax_ref = cmax, auc_ref = aucinf.obs,
                              thalf_ref = half.life, tmax_ref = tmax),
    by = "arm"
  )

auc_pct   <- 100 * (nca_wide$aucinf.obs - nca_wide$auc_ref)   / nca_wide$auc_ref
cmax_pct  <- 100 * (nca_wide$cmax       - nca_wide$cmax_ref)  / nca_wide$cmax_ref
thalf_pct <- 100 * (nca_wide$half.life  - nca_wide$thalf_ref) / nca_wide$thalf_ref

# Realised: AUC 0.001%, Cmax 0.000%, t1/2 -0.62% (NLME) / -0.26% (NLLS).
# AUC and Cmax are exact identities, so they are held to 0.05%. The t1/2 bound
# is looser because PKNCA selects its own lambda-z window (from 10.5 h here),
# which still carries a trace of the alpha phase; 2% admits that and still
# goes red on a mis-transcribed Q or V2, either of which moves beta by tens of
# percent.
stopifnot(
  max(abs(auc_pct))   < 0.05,
  max(abs(cmax_pct))  < 0.05,
  max(abs(thalf_pct)) < 2
)
c(auc = signif(max(abs(auc_pct)), 3),
  cmax = signif(max(abs(cmax_pct)), 3),
  half_life = signif(max(abs(thalf_pct)), 3))
```

`AUC(0-inf)` also has to be exactly proportional to dose, since neither model
has any nonlinearity:

```{r nca-linearity}
lin <- nca_wide |>
  dplyr::mutate(dose_mg = as.numeric(sub("^\\w+ (\\d+) mg$", "\\1", arm)),
                model   = sub("^(\\w+) .*$", "\\1", arm),
                auc_per_mg = aucinf.obs / dose_mg) |>
  dplyr::group_by(model) |>
  dplyr::summarise(spread_pct = 100 * (max(auc_per_mg) - min(auc_per_mg)) /
                     mean(auc_per_mg), .groups = "drop")
stopifnot(all(lin$spread_pct < 0.01))
lin
```

# Reproducing the paper's central quantitative claims

Each row below restates a claim the paper makes in prose and checks it against
the packaged models. Every quantity is deterministic.

```{r claims}
p_nlme <- pars$NLME
p_nlls <- pars$NLLS

# Paper's "Literature [15]" column of Table 4, transcribed. Independently
# available in this library as Cojutti_2024_dalbavancin.
ui_coj  <- rxode2::rxode(readModelDb("Cojutti_2024_dalbavancin"))
coj_th  <- exp(ui_coj$theta[c("lcl", "lvc", "lq", "lvp")])
lit_col <- c(lcl = 0.031, lvc = 5.9, lq = 0.038, lvp = 9.6)
lit_pct <- 100 * (coj_th - lit_col) / lit_col

# Time above the 8.04 mg/L efficacy threshold after one 1500 mg infusion.
t_above <- sim |>
  dplyr::filter(dose_mg == 1500, time > T_INF, Cc >= 8.04) |>
  dplyr::group_by(model) |>
  dplyr::summarise(t_last = max(time), .groups = "drop")
t_above_nlme <- t_above$t_last[t_above$model == "NLME"]
t_above_nlls <- t_above$t_last[t_above$model == "NLLS"]

# V_NODE, Equation 5, as an independent check on V1.
v_node     <- 5.7
v_node_pct <- 100 * (p_nlme$vc - v_node) / v_node

claim <- function(text, source, achieved, pass) {
  data.frame(Claim = text, Source = source, Achieved = achieved, Pass = pass)
}

claims <- dplyr::bind_rows(
  claim(
    "The two-compartment (NLLS) fit fails to estimate Q accurately",
    "Discussion",
    sprintf("Q = %.3f L/h, %.1fx the NLME estimate and %.1fx the literature value",
            p_nlls$q, p_nlls$q / p_nlme$q, p_nlls$q / 0.038),
    p_nlls$q / p_nlme$q > 5
  ),
  claim(
    "The two-compartment fit systematically underestimates long-term concentrations",
    "Figure 4, Discussion",
    sprintf("Cc at 28 d after 1500 mg: %.2f mg/L (NLLS) vs %.2f mg/L (NLME), %.0f%% lower",
            sim$Cc[sim$model == "NLLS" & sim$dose_mg == 1500 & sim$time == 672],
            sim$Cc[sim$model == "NLME" & sim$dose_mg == 1500 & sim$time == 672],
            100 * (1 - sim$Cc[sim$model == "NLLS" & sim$dose_mg == 1500 & sim$time == 672] /
                     sim$Cc[sim$model == "NLME" & sim$dose_mg == 1500 & sim$time == 672])),
    sim$Cc[sim$model == "NLLS" & sim$dose_mg == 1500 & sim$time == 672] <
      sim$Cc[sim$model == "NLME" & sim$dose_mg == 1500 & sim$time == 672]
  ),
  claim(
    "... with a materially shorter terminal half-life",
    "derived from Table 4",
    sprintf("t1/2 beta = %.0f h (%.1f d) NLLS vs %.0f h (%.1f d) NLME",
            p_nlls$t_half_beta, p_nlls$t_half_beta / 24,
            p_nlme$t_half_beta, p_nlme$t_half_beta / 24),
    p_nlls$t_half_beta < 0.5 * p_nlme$t_half_beta
  ),
  claim(
    "... which changes the predicted duration above the 8.04 mg/L target",
    "Methods 2.1 threshold; Figure 3",
    sprintf("%.0f d (NLLS) vs %.0f d (NLME) after a single 1500 mg dose",
            t_above_nlls / 24, t_above_nlme / 24),
    t_above_nlls < t_above_nlme
  ),
  claim(
    "The NLME estimates are consistent with previous literature estimates",
    "Discussion; Table 4 Literature column",
    sprintf("Vss %.1f L vs %.1f L (%+.0f%%); CL %.4f vs %.4f L/h (%+.0f%%)",
            p_nlme$vss, 5.9 + 9.6, 100 * (p_nlme$vss / (5.9 + 9.6) - 1),
            p_nlme$cl, 0.031, 100 * (p_nlme$cl / 0.031 - 1)),
    abs(100 * (p_nlme$vss / (5.9 + 9.6) - 1)) < 50
  ),
  claim(
    "The NODE volume V_NODE assumes values similar to V1",
    "Equation 5; Discussion",
    sprintf("V1 = %.2f L vs V_NODE = %.1f L (%+.1f%%)", p_nlme$vc, v_node, v_node_pct),
    abs(v_node_pct) < 20
  ),
  claim(
    "This library's Cojutti 2024 extraction matches the paper's Literature column",
    "Table 4 Literature column vs Cojutti_2024_dalbavancin",
    sprintf("max deviation %.2f%% across CL, V1, Q, V2", max(abs(lit_pct))),
    max(abs(lit_pct)) < 1
  )
)

stopifnot(all(claims$Pass))

knitr::kable(claims, caption = "Published claims checked against the packaged models.")
```

The last row is a cross-extraction consistency check rather than a claim of
this paper: Giacometti 2025 transcribes Cojutti 2024's estimates into Table 4's
"Literature" column, and this library already carries Cojutti 2024 as an
independent extraction. The two agree to rounding, which validates both
transcriptions.

## Derived disposition summary

```{r summary-table}
tibble::tibble(
  Estimation = c("NLLS (naive pooled)", "NLME (Monolix)", "Cojutti 2024 (library)"),
  `CL (L/h)` = c(p_nlls$cl, p_nlme$cl, coj_th[["lcl"]]),
  `Q (L/h)`  = c(p_nlls$q,  p_nlme$q,  coj_th[["lq"]]),
  `V1 (L)`   = c(p_nlls$vc, p_nlme$vc, coj_th[["lvc"]]),
  `V2 (L)`   = c(p_nlls$vp, p_nlme$vp, coj_th[["lvp"]]),
  `Vss (L)`  = c(p_nlls$vss, p_nlme$vss, coj_th[["lvc"]] + coj_th[["lvp"]]),
  `t1/2 alpha (h)` = c(log(2) / p_nlls$alpha, log(2) / p_nlme$alpha, NA),
  `t1/2 beta (d)`  = c(p_nlls$t_half_beta / 24, p_nlme$t_half_beta / 24, NA)
) |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) signif(x, 4))) |>
  knitr::kable(caption = "Disposition parameters implied by each fit.")
```

# Assumptions and deviations

* **Inter-individual variability is not reported for the NLME arm, and is
  encoded as absent.** This is a genuine mixed-effects fit -- Methods 2.2
  describes the random-effects framework and states that Monolix estimated the
  model by maximum likelihood -- so omegas were certainly estimated. But
  Table 4 is the paper's only parameter table and it prints four rows (CL, Q,
  V1, V2) with no omega block, no shrinkage and no CV%. **The plus-or-minus
  figures in Table 4 are the standard deviation across the six
  cross-validation folds (Table 4 caption), not variance components**, and
  using them as omegas would be a category error: the NLME CL spread of
  0.0006 L/h on 0.0367 L/h is a 1.6% across-fold stability figure, which is
  implausibly small for between-subject variability in a TDM cohort spanning
  serum creatinine 19 to 411 umol/L. No variance is invented. No `eta` is
  declared at all -- rather than `etalcl ~ fixed(0)` -- because a zero-variance
  diagonal makes OMEGA singular and breaks the Cholesky sampler in `rxSolve`.
  Simulations from both files are typical-value only.
* **For the NLLS arm the absence is structural, not a reporting gap.** A
  naive-pooled least-squares fit has no random-effects layer: one curve is
  fitted through the pooled observations of all 218 patients.
* **Residual error is not reported for either arm.** Methods 2.2 gives the
  objective as a plain squared loss on observed minus predicted concentration,
  but no residual SD, %CV or sigma appears anywhere in the article. `propSd`
  and `addSd` are therefore both `fixed(0)`. Both a proportional and an
  additive term are carried so a user with an external estimate can supply
  either through `ini()` without editing `model()`.
* **The Neural ODE arm is not packaged.** Its several hundred trained network
  weights are not published, and a layered network cannot be expressed inside
  an rxode2 `model()` block. Only `V_NODE = 5.7 +/- 0.1 L` (Equation 5) and
  `V_NODE,cov = 5.2 +/- 0.8 L` (Equation 6) are reported numerically, and
  `V_NODE` is used above only as an independent cross-check on `V1`. This is
  the paper's headline model, so users should read the two packaged files as
  the paper's comparators rather than as its conclusion.
* **Infusion duration is assumed to be 30 minutes.** Methods 2.2 specifies
  "infusion, no delay" but the article never gives a duration, and the
  sibling dalbavancin extractions note the same gap. 30 min is the
  dalbavancin label duration. The assumption affects only `Cmax` and the first
  minutes of the profile; every check in this vignette other than `Cmax` is
  invariant to it, and `Cmax` is compared against the analytical solution at
  the same assumed duration rather than against a published value.
* **The dose levels simulated here (350, 500, 1000, 1500 mg) span the range
  the paper reports (Methods 2.1) but are not its actual regimens.** The
  article gives neither the dosing intervals nor the per-patient schedules,
  so single-dose profiles are used throughout.
* **Equation 2a is read on amounts, not concentrations.** As printed and read
  literally on concentrations it does not conserve mass; on amounts it is the
  canonical two-compartment system, which is also what Monolix was configured
  with and what Figure 2 depicts. See "Reading Equation 2a" above; the
  mass-balance check is the mechanical confirmation.
* **Serum creatinine is reported as "creatine" throughout the source.**
  Table 1's row label is "Creatine conc. (umol/L)" and Methods 2.1 says
  "serum creatine concentration", but the reported values (mean 94, range 19
  to 411 umol/L; 94 umol/L is 1.06 mg/dL) are a normal serum creatinine and
  far outside any plausible plasma creatine concentration. Recorded as
  canonical `CREAT` in `covariatesDataExcluded`. The covariate is not used by
  either model, so nothing downstream depends on the reading.
* **No covariate is carried, by the authors' design.** Age, height, weight,
  sex and serum creatinine were all collected and screened; Appendix B
  reports the covariate-free model was kept (Kolmogorov-Smirnov p = 0.45
  against the per-fold best covariate model). This contrasts with the two
  sibling dalbavancin models in this library, which do carry covariates:
  `Cojutti_2024_dalbavancin` retains CKD-EPI eGFR on CL, and
  `Baiardi_2025_dalbavancin` applies fixed-exponent allometric weight
  scaling. All five screened covariates are preserved in
  `covariatesDataExcluded`.
* **Race and ethnicity are not reported** by the source; the cohort is
  single-centre Italian.
* **All values come from the article's text and tables.** No supplement
  accompanies this paper; the linked GitHub repository holds analysis code
  only, and no value in either model file is taken from it.
