---
title: "Evolocumab (Gibbs 2017)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evolocumab (Gibbs 2017)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Gibbs_2017_evolocumab"))$reference`
* Description: `r rxode2::rxode(readModelDb("Gibbs_2017_evolocumab"))$description`
* Article: <https://doi.org/10.1002/jcph.840>

Evolocumab is a fully human IgG2 monoclonal antibody against proprotein
convertase subtilisin/kexin type 9 (PCSK9). PCSK9 promotes degradation of the
low-density lipoprotein receptor, so removing unbound PCSK9 increases LDL
receptor recycling and accelerates clearance of LDL-cholesterol (LDL-C).
Because PCSK9 is an abundant, high-turnover soluble target, binding to it is
also a *route of elimination* for evolocumab itself, which is what makes the
drug's pharmacokinetics nonlinear.

Gibbs 2017 links three layers in one simultaneously-fitted model:

1. a one-compartment evolocumab PK model with first-order subcutaneous (SC)
   absorption and **linear plus target-mediated** elimination, using the
   quasi-steady-state (QSS) approximation of the full TMDD model;
2. a **PCSK9 turnover pool** whose degradation is accelerated by binding to
   evolocumab and elimination of the complex; and
3. a **type-4 indirect-response model** for LDL-C in which unbound PCSK9
   inhibits LDL-C elimination.

The paper's purpose is a dosing question that the coupling makes
counter-intuitive: doubling the dose does **not** buy a doubled dosing
interval. Both the target-mediated PK nonlinearity and the saturable
PCSK9 -> LDL-C relationship compress the benefit of a larger dose, so an
approximately 3-fold dose increase (140 mg Q2W -> 420 mg QM) is needed to hold
the same time-averaged LDL-C reduction over a monthly interval.

# Population

The model was fitted to pooled data from two phase 1 studies in 101
individuals contributing 4910 observations of unbound evolocumab, unbound
PCSK9, and LDL-C:

* **Study 20080397 (phase 1a)** -- 44 healthy subjects, single ascending SC
  doses of 7, 21, 70, 210, or 420 mg.
* **Study 20080398 (phase 1b)** -- 57 patients with hypercholesterolemia on
  stable low- to moderate-intensity statin therapy, multiple ascending SC
  doses of 14 mg QW x 6, 35 mg QW x 6, 140 mg Q2W x 3, 280 mg Q2W x 3, or
  420 mg QM x 2. Two further cohorts received high-intensity statin therapy or
  carried a heterozygous familial hypercholesterolemia diagnosis and were dosed
  140 mg Q2W x 3.

Across both studies 73 participants received evolocumab and 28 received
placebo. Most were male (73.3%) and white (83.2%), with a mean age of 45.5
years and a mean body weight of 81.4 kg. Baseline PCSK9 was 1.58-fold higher in
the phase 1b study than in the phase 1a study, because statins upregulate
PCSK9 -- the single covariate the final model retains.

Estimation used NONMEM 7.2 (SAEM followed by importance sampling), with the M3
method for below-quantification-limit unbound evolocumab and unbound PCSK9
observations.

The same information is available programmatically via the model's
`population` metadata:

```{r population}
str(readModelDb("Gibbs_2017_evolocumab")()$population)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Gibbs_2017_evolocumab.R`. The table below
collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `d/dt(depot)` | n/a | Equation 1 |
| `d/dt(central)` (total drug amount, TDA) | n/a | Equation 2 |
| `d/dt(total_target)` (total ligand conc., TLC) | n/a | Equation 3 |
| `tdc <- central / vc` | n/a | Equation 4 |
| `fdc` (free drug conc., QSS quadratic root) | n/a | Equation 5 |
| `flc <- tlc - (tdc - fdc)` | n/a | Equation 6 |
| `d/dt(ldl)` | n/a | Equation 7 |
| `ksyn <- kdeg * rbase_target` | n/a | Methods: "related to the baseline PCSK9 (BASE_PCSK9) level as ksyn/kdeg" |
| `lka` | 0.245 1/day | Table 2 |
| `lcl` | 0.256 L/day | Table 2 |
| `lvc` | 2.66 L | Table 2 |
| `lfdepot` | 0.72 (fixed) | Methods: "Bioavailability (F) was fixed to 72% (Amgen internal data)" |
| `lkdeg` | 2.12 1/day | Table 2 |
| `lrbase_target` | 5.27 nM | Table 2; Results ("5.27 nM (or 379 ng/mL)") |
| `lkss` | 0.253 nM | Table 2 |
| `lkint` | 0.0529 1/day | Table 2 |
| `e_dis_healthy_rbase_target` | 0.637 | Table 2 (theta 1); Results ("3.36 nM (or 242 ng/mL) in healthy subjects") |
| `lkout` | 0.305 1/day | Table 3 |
| `lrbase_ldl` | 116 mg/dL | Table 3 |
| `limax` | 1.00 (fixed) | Table 3; Results ("near the boundary of 1.00, so its value was fixed") |
| `lic50` | 1.46 nM | Table 3; Abstract |
| `etalka + etalvc + etalcl` block | 0.807 / 0.158 / 0.924; cov 0.253 / -0.671 / -0.183 | Table 2 random-effects and Covariance rows |
| `etalrbase_target` | 0.0348 | Table 2 |
| `etalkss` | 1.17 | Table 2 |
| `etalrbase_ldl` | 0.0465 | Table 3 |
| `etalic50` | 0.481 | Table 3 |
| `propSd` (unbound evolocumab) | sqrt(0.0576) = 0.240 | Table 2 sigma^2 (evolocumab) |
| `propSd_freePcsk9` | sqrt(0.0942) = 0.3069 | Table 2 sigma^2 (PCSK9) |
| `propSd_ldl` | sqrt(0.0130) = 0.1140 | Table 3 sigma^2 (proportional) |
| `addSd_ldl` | sqrt(60.1) = 7.7524 mg/dL | Table 3 sigma^2 (additive), footnote b |
| `MW_EVO` = 141800 g/mol | -- | **Not in the paper** -- see Assumptions and deviations |

## Reading the random-effects columns

Both tables print the random-effects column as `Estimate (~CV%)`, and the
footnote defines "CV%, coefficient of variation calculated as sqrt(var) x 100".
The estimates are therefore **variances**, carried into `ini()` unchanged. Each
printed CV% reproduces from the estimate, which confirms the reading rather
than assuming it:

```{r variance-check}
omegas <- c(ka = 0.807, V = 0.158, CL = 0.924, BASE_PCSK9 = 0.0348,
            kss = 1.17, BASE_LDLC = 0.0465, IC50 = 0.481)
published_cv <- c(90, 40, 96, 19, 108, 22, 69)
round(sqrt(omegas) * 100)
stopifnot(abs(round(sqrt(omegas) * 100) - published_cv) <= 1)
```

The parenthesised percentages on the three `Covariance` rows are correlations,
and they too reproduce from the covariances -- confirming the ka/V/CL block is
a covariance (not a correlation) matrix, and that it is positive definite:

```{r covariance-check}
blk <- matrix(c( 0.807,  0.253, -0.671,
                 0.253,  0.158, -0.183,
                -0.671, -0.183,  0.924), 3, 3,
              dimnames = list(c("ka", "V", "CL"), c("ka", "V", "CL")))
round(stats::cov2cor(blk) * 100)      # published: 70, -78, -48
stopifnot(
  abs(round(stats::cov2cor(blk)[1, 2] * 100) -  70) <= 1,
  abs(round(stats::cov2cor(blk)[1, 3] * 100) - -78) <= 1,
  abs(round(stats::cov2cor(blk)[2, 3] * 100) - -48) <= 1,
  # Positive definite, so rxode2's Cholesky sampler can draw from it as-is
  # (no off-diagonal nudge needed; see known-vignette-failure-patterns.md #1).
  all(eigen(blk, symmetric = TRUE, only.values = TRUE)$values > 0)
)
```

# Model structure

```{r model}
mod <- readModelDb("Gibbs_2017_evolocumab")
mod
```

Equation 5 is the positive root of the one-to-one QSS binding quadratic. With
`RC` the complex concentration, QSS binding gives `RC = FDC * FLC / kss`, and
mass balance gives `FLC = TLC - RC` and `RC = TDC - FDC`; eliminating `RC` and
`FLC` yields

```
FDC^2 + FDC * (TLC - TDC + kss) - kss * TDC = 0
```

whose positive root is the printed equation 5. The PDF text layer loses the
radical and the squared term of that expression, so it was re-derived here and
checked against the printed `4 kss TDC` term, which it reproduces exactly.

A useful consequence, used repeatedly below: with `Imax` fixed at 1, the
maximum attainable LDL-C reduction (complete PCSK9 suppression) is
`BASE_PCSK9 / (IC50 + BASE_PCSK9)`.

```{r ceiling}
max_reduction <- 5.27 / (1.46 + 5.27)
round(100 * max_reduction, 1)
```

# Typical-value replication

Gibbs 2017 performed its dose-regimen simulations deterministically -- "we
performed simulations using the typical values of PK/PD parameters using
Berkley Madonna" -- so the replications below use `rxode2::zeroRe()` to match
that. The stochastic cohort comes afterwards.

```{r typical-model}
mod_typ <- rxode2::zeroRe(mod)

# This is a three-endpoint model (Cc, freePcsk9, ldl), so every event table
# below is built as an explicit data frame with `evid`, `cmt` and `dvid`
# columns: an observation row that names no compartment cannot be mapped to an
# endpoint and rxSolve rejects it ("'dvid'->'cmt' ... on observation record").
# Observations are placed on the ODE state `central` -- never on an algebraic
# observable name such as "Cc", which would inject an extra compartment slot
# and renumber the states -- and every model variable is returned as its own
# column regardless of which endpoint `dvid` names. `useLinCmt = FALSE` is
# required for the same reason (see known-vignette-failure-patterns.md #2, #5b).
make_events <- function(amt, ii, addl, until, dis_healthy = 0,
                        by = 0.05, id = 1L) {
  obs <- data.frame(id = id, time = seq(0, until, by = by), amt = NA_real_,
                    evid = 0L, cmt = "central", dvid = 1L,
                    DIS_HEALTHY = dis_healthy)
  if (is.null(amt) || amt <= 0) return(obs[order(obs$time), ])
  dose <- data.frame(id = id, time = ii * seq(0, addl), amt = amt, evid = 1L,
                     cmt = "depot", dvid = NA_integer_,
                     DIS_HEALTHY = dis_healthy)
  ev <- rbind(dose, obs)
  ev[order(ev$time, -ev$evid), ]
}

# Solve one regimen for one baseline-PCSK9 state.
# DIS_HEALTHY = 0 is the statin-treated patient (BASE_PCSK9 = 5.27 nM).
solve_regimen <- function(amt, ii, addl, until, dis_healthy = 0,
                          model = mod_typ, by = 0.05) {
  rxode2::rxSolve(
    model, make_events(amt, ii, addl, until, dis_healthy, by = by),
    useLinCmt = FALSE, returnType = "data.frame"
  )
}
```

## Baseline holds exactly with no drug

Both turnover pools are initialised at their drug-free steady states
(`ksyn = kdeg * BASE_PCSK9`, and `kin` solved from the LDL-C balance), so an
undosed solve must be flat. This is a pure numerical identity -- the two sides
share the same parameters -- so a tight bound is the correct assertion here.

```{r baseline-hold}
hold <- solve_regimen(amt = 0, ii = 0, addl = 0, until = 120)
c(max_ldl_drift   = max(abs(hold$ldl - 116)),
  max_pcsk9_drift = max(abs(hold$total_target - 5.27)))
stopifnot(
  max(abs(hold$ldl - 116)) < 1e-6,
  max(abs(hold$total_target - 5.27)) < 1e-6,
  # Free and total PCSK9 coincide when no drug is present (equation 6 with TDC = 0).
  max(abs(hold$freePcsk9 - hold$total_target)) < 1e-9
)
```

## The healthy-subject covariate reproduces the paper's second baseline

The paper prints both typical baselines, which over-determines the coding of
`DIS_HEALTHY` and fixes its direction.

```{r healthy-baseline}
hold_hv <- solve_regimen(amt = 0, ii = 0, addl = 0, until = 10, dis_healthy = 1)
c(patient_nM = hold$total_target[1], healthy_nM = hold_hv$total_target[1])
stopifnot(
  abs(hold$total_target[1]    - 5.27) < 1e-6,   # Results: 5.27 nM (379 ng/mL)
  abs(hold_hv$total_target[1] - 3.36) < 0.01    # Results: 3.36 nM (242 ng/mL)
)
```

The paper's ng/mL <-> nM conversions are internally consistent and imply a
PCSK9 molecular weight of about 72 kDa (379/5.27 = 71.9; 242/3.36 = 72.0),
which is used below wherever PCSK9 must be expressed in ng/mL.

```{r pcsk9-mw}
mw_pcsk9 <- c(patients = 379 / 5.27, healthy = 242 / 3.36)   # kDa
round(mw_pcsk9, 1)
stopifnot(abs(diff(mw_pcsk9)) < 0.5)

mw_pcsk9_kda <- mean(mw_pcsk9)   # used wherever PCSK9 must cross ng/mL <-> nM
```

## Figure 5 -- LDL-C time course after multiple SC doses

Replicates Figure 5 of Gibbs 2017: the time course of LDL-C over 12 weeks in
statin-treated patients (BASE_PCSK9 = 5.27 nM) at 140 mg SC Q2W, 280 mg SC QM,
and 420 mg SC QM.

```{r figure-5}
arms <- tibble::tribble(
  ~regimen,       ~amt, ~ii, ~addl,
  "140 mg Q2W",    140,  14,     5,   # doses at weeks 0, 2, 4, 6, 8, 10
  "280 mg QM",     280,  28,     2,   # doses at weeks 0, 4, 8
  "420 mg QM",     420,  28,     2
)

fig5 <- do.call(rbind, lapply(seq_len(nrow(arms)), function(i) {
  s <- solve_regimen(arms$amt[i], arms$ii[i], arms$addl[i], until = 84)
  data.frame(regimen = arms$regimen[i], time = s$time, ldl = s$ldl,
             freePcsk9 = s$freePcsk9, Cc = s$Cc)
}))

ggplot(fig5, aes(time / 7, ldl, colour = regimen)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 116, linetype = "dotted") +
  geom_hline(yintercept = 116 * (1 - max_reduction), linetype = "dashed") +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  labs(x = "Time (weeks)", y = "LDL-C (mg/dL)", colour = NULL,
       title = "Figure 5 - LDL-C after multiple SC evolocumab doses",
       caption = paste("Replicates Figure 5 of Gibbs 2017. Dotted = baseline 116 mg/dL;",
                       "dashed = complete-PCSK9-suppression floor.")) +
  theme_bw() + theme(legend.position = "top")
```

## Time-averaged LDL-C reduction (the paper's headline numbers)

Gibbs 2017 reports the time-averaged effect over the steady-state window
(weeks 8 to 12) as **68.9%** for 140 mg Q2W, **63.5%** for 280 mg QM, and
**68.9%** for 420 mg QM -- the quantitative basis for its "3-fold dose increase
for a 2-fold interval extension" conclusion.

```{r tavg}
tavg <- fig5 |>
  filter(time >= 56, time <= 84) |>
  group_by(regimen) |>
  summarise(
    simulated_pct = mean(100 * (1 - ldl / 116)),
    trough_pct    = min(100 * (1 - ldl / 116)),
    nadir_pct     = max(100 * (1 - ldl / 116)),
    .groups = "drop"
  ) |>
  left_join(
    tibble::tibble(regimen  = c("140 mg Q2W", "280 mg QM", "420 mg QM"),
                   published_pct = c(68.9, 63.5, 68.9)),
    by = "regimen"
  ) |>
  mutate(difference_pp = simulated_pct - published_pct)

tavg |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  dplyr::rename(
    "Regimen"                          = regimen,
    "Simulated time-averaged (%)"      = simulated_pct,
    "Trough (%)"                       = trough_pct,
    "Nadir (%)"                        = nadir_pct,
    "Published (%)"                    = published_pct,
    "Difference (percentage points)"   = difference_pp
  ) |>
  knitr::kable(
    caption = paste("Time-averaged LDL-C reduction over weeks 8-12 versus the values",
                    "reported in Gibbs 2017 Results.")
  )
```

The reproduction is within about 2 percentage points on all three arms, and
reproduces the paper's two structural conclusions exactly: 280 mg QM is the
clear loser, and 140 mg Q2W and 420 mg QM are equivalent.

These assertions are safe to keep tight because this whole section is
deterministic -- `zeroRe()` removes every random effect, so there is no cohort
draw to vary across rxode2 builds or CI thread counts.

```{r tavg-assertions}
stopifnot(
  # Each arm reproduces its published value. Observed differences at authoring:
  # +1.4, +2.0, +2.0 percentage points; the bound leaves headroom but a
  # mis-transcribed CL, IC50, dose or unit conversion moves these by tens of points.
  all(abs(tavg$difference_pp) < 3.5),
  # Conclusion 1: 280 mg QM is materially worse than either of the other arms.
  # Observed gaps at authoring: 4.8 and 5.4 percentage points; published: 5.4 both.
  with(tavg, simulated_pct[regimen == "140 mg Q2W"] -
             simulated_pct[regimen == "280 mg QM"]) > 3,
  with(tavg, simulated_pct[regimen == "420 mg QM"] -
             simulated_pct[regimen == "280 mg QM"]) > 3,
  # Conclusion 2: 140 mg Q2W and 420 mg QM are equivalent (published: both 68.9%).
  # Observed gap at authoring: 0.7 percentage points.
  with(tavg, abs(simulated_pct[regimen == "140 mg Q2W"] -
                 simulated_pct[regimen == "420 mg QM"])) < 2,
  # Conclusion 3: 280 mg QM fails by fluctuating over the interval, not by a
  # lower peak effect -- its nadir is comparable while its trough is far worse.
  with(tavg, trough_pct[regimen == "280 mg QM"] <
             trough_pct[regimen == "140 mg Q2W"] - 15)
)
```

## Unbound PCSK9 suppression and its duration

The paper makes two separate observational claims about the single-dose study:
that doses of 21 mg and above produced "> 80% suppression of unbound PCSK9",
and that "the duration of **full** suppression" rose with dose "up to a maximum
of 14 days" at 420 mg. The two are different thresholds -- *full* suppression
means unbound PCSK9 driven below the assay's lower limit of quantification
(15 ng/mL), which from a 3.36 nM healthy baseline is a 93.8% reduction.

```{r pcsk9-suppression}
lloq_pcsk9_nM <- 15 / mw_pcsk9_kda   # 15 ng/mL assay LLOQ -> nM

single_doses <- c(7, 21, 70, 210, 420)
supp <- do.call(rbind, lapply(single_doses, function(d) {
  s <- solve_regimen(d, ii = 0, addl = 0, until = 90, dis_healthy = 1)
  base_hv <- s$total_target[1]
  blq <- s$freePcsk9 < lloq_pcsk9_nM
  data.frame(
    dose_mg         = d,
    max_suppression = 100 * (1 - min(s$freePcsk9) / base_hv),
    days_full_supp  = if (any(blq)) max(s$time[blq]) else 0
  )
}))
supp |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  dplyr::rename("Single SC dose (mg)"                   = dose_mg,
                "Maximum PCSK9 suppression (%)"         = max_suppression,
                "Days of full suppression (< 15 ng/mL)" = days_full_supp) |>
  knitr::kable(caption = paste("Unbound PCSK9 suppression after a single SC dose in",
                               "healthy subjects (typical-value prediction)."))
```

The dose-ordering and the depth of suppression reproduce, and the 7 mg dose
correctly fails to suppress -- consistent with the paper's report that unbound
evolocumab was below the detection limit in every healthy subject after 7 mg.
Two quantitative details do **not** reproduce exactly and are recorded as known
deviations rather than gated: the typical-value prediction reaches 76%
suppression at 21 mg where the paper reports "> 80%", and it holds full
suppression for about 21 days at 420 mg where the paper reports a maximum of
14 days. Both published numbers summarise *observed* data read off a discrete
visit schedule (days 1, 3, 4, 5, 6, 7, 8, 11, 15, 22, 29, ...), so their
resolution near 14 days is roughly a week, and both are individual-level
observations rather than typical-value predictions. See Assumptions and
deviations.

```{r pcsk9-assertions}
stopifnot(
  # Deep suppression at the doses where the paper's claim is unambiguous.
  all(supp$max_suppression[supp$dose_mg >= 70] > 90),
  # The 7 mg dose does not meaningfully suppress; 21 mg clearly does.
  supp$max_suppression[supp$dose_mg == 7]  < 50,
  supp$max_suppression[supp$dose_mg == 21] > 60,
  # Both metrics rise monotonically with dose (the paper's stated trend).
  all(diff(supp$max_suppression) > 0),
  all(diff(supp$days_full_supp) >= 0),
  # Full suppression at 420 mg lasts on the order of the published 14 days
  # (deterministic, so this window is a real gate, not a formality).
  supp$days_full_supp[supp$dose_mg == 420] > 7,
  supp$days_full_supp[supp$dose_mg == 420] < 35
)
```

# Virtual cohort and stochastic simulation

The section above is deterministic. To exercise the published between-subject
variability -- which is large (CV 96% on CL, 108% on kss) -- the cohort below
simulates the phase 1b 140 mg Q2W x 3 regimen with all random effects active.

```{r cohort}
# `set.seed()` seeds R's RNG, NOT rxode2's simulation RNG, and rxode2's streams
# are partitioned per solver thread -- so this cohort is reproducible on this
# machine and different on a machine with a different thread count. Every
# assertion below is therefore written on medians and robust quantiles, never
# on a cohort extreme.
set.seed(20170840)
rxode2::rxSetSeed(20170840)

n_per_arm <- 200L   # cohort cap is 200 per arm

make_cohort <- function(n, dis_healthy, label, id_offset = 0L) {
  do.call(rbind, lapply(seq_len(n), function(i) {
    ev <- make_events(amt = 140, ii = 14, addl = 2, until = 85,
                      dis_healthy = dis_healthy, by = 0.5,
                      id = id_offset + i)
    ev$cohort <- label
    ev
  }))
}

events <- rbind(
  make_cohort(n_per_arm, 0, "Statin-treated patients", id_offset =   0L),
  make_cohort(n_per_arm, 1, "Healthy subjects",        id_offset = 200L)
)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

```{r simulate}
sim <- rxode2::rxSolve(
  mod, events = events,
  keep = c("cohort", "DIS_HEALTHY"),
  useLinCmt = FALSE
) |> as.data.frame()

# The QSS free-drug root involves a square root; confirm the solve produced no
# NaN anywhere before any of it is summarised.
stopifnot(
  all(is.finite(sim$Cc)), all(is.finite(sim$freePcsk9)), all(is.finite(sim$ldl)),
  all(sim$Cc >= -1e-8), all(sim$freePcsk9 >= -1e-8), all(sim$ldl > 0),
  # Free ligand can never exceed total ligand (equations 5 and 6).
  all(sim$freePcsk9 <= sim$total_target + 1e-8)
)
```

## Figure 4B -- visual predictive check, 140 mg SC Q2W x 3

Replicates Figure 4B of Gibbs 2017: median and 90% prediction interval for
unbound evolocumab, unbound PCSK9, and LDL-C in the statin-treated cohort.

```{r figure-4b}
vpc <- sim |>
  filter(cohort == "Statin-treated patients") |>
  select(time, Cc, freePcsk9, ldl) |>
  pivot_longer(-time, names_to = "analyte", values_to = "value") |>
  mutate(analyte = factor(
    analyte,
    levels = c("Cc", "freePcsk9", "ldl"),
    labels = c("Unbound evolocumab (nM)", "Unbound PCSK9 (nM)", "LDL-C (mg/dL)")
  )) |>
  group_by(analyte, time) |>
  summarise(Q05 = quantile(value, 0.05), Q50 = median(value),
            Q95 = quantile(value, 0.95), .groups = "drop")

ggplot(vpc, aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(linewidth = 0.7) +
  facet_wrap(~analyte, scales = "free_y") +
  labs(x = "Time (days)", y = NULL,
       title = "Figure 4B - VPC, 140 mg SC evolocumab Q2W x 3",
       caption = "Replicates Figure 4B of Gibbs 2017 (median and 90% prediction interval).") +
  theme_bw()
```

```{r vpc-assertions}
# Robust, median-based checks only -- these must hold for ANY cohort the model
# can draw, so nothing here depends on which subjects landed in the tails.
ldl_med <- vpc |>
  filter(analyte == "LDL-C (mg/dL)", time >= 28, time <= 42) |>
  summarise(reduction = mean(100 * (1 - Q50 / 116))) |>
  pull(reduction)

stopifnot(
  # The median subject's LDL-C reduction after 3 doses sits in the same region
  # as the deterministic prediction, well short of the structural ceiling.
  ldl_med > 40, ldl_med < 100 * max_reduction + 1,
  # PCSK9 is deeply suppressed at the median right after a dose ...
  min(vpc$Q50[vpc$analyte == "Unbound PCSK9 (nM)"]) < 1,
  # ... and the median recovers toward, but does not overshoot, baseline by the
  # end of follow-up. Paper: "returned toward the baseline without evidence of
  # rebound above the initial baseline values."
  max(vpc$Q50[vpc$analyte == "Unbound PCSK9 (nM)"]) < 5.27 * 1.05
)
```

## Statin intensity as a target-mediated-elimination falsifier

Gibbs 2017 Table 1 reports a within-study contrast that is a strong,
schedule-independent check on the whole TMDD layer: among patients given
140 mg Q2W, those on **high-intensity** statins had 26% higher baseline PCSK9
(487 vs 385 ng/mL) and, consequently, **20% lower** unbound evolocumab Cmax and
AUClast (ratios 0.80 and 0.80). A model with a mis-specified target sink would
not reproduce that coupling.

The two strata differ only in baseline PCSK9, so they are simulated by setting
`lrbase_target` to each stratum's observed value rather than by a covariate.

```{r statin-strata}
strata <- tibble::tibble(
  treatment      = c("Low-moderate intensity statin", "High intensity statin"),
  pcsk9_ng_ml    = c(385, 487),
  base_nM        = c(385, 487) / mw_pcsk9_kda   # paper's own ng/mL <-> nM bridge
)

mw_evo <- 141800   # g/mol; see Assumptions and deviations
lloq_ug_ml <- 0.8  # assay lower limit of quantification (Methods)

strata_sim <- do.call(rbind, lapply(seq_len(nrow(strata)), function(i) {
  m <- rxode2::ini(mod_typ, lrbase_target = log(strata$base_nM[i]))
  s <- rxode2::rxSolve(m, make_events(amt = 140, ii = 14, addl = 2, until = 85,
                                      dis_healthy = 0, by = 0.25),
                       useLinCmt = FALSE, returnType = "data.frame")
  data.frame(id = i, treatment = strata$treatment[i], time = s$time,
             Cc = s$Cc * mw_evo / 1e6)   # nM -> ug/mL, the paper's Table 1 unit
}))
```

# PKNCA validation

NCA is run on the typical-value profiles, matching how Gibbs 2017 generated its
own simulated summaries. Concentrations are truncated at the assay LLOQ so that
`auclast` reproduces the paper's definition -- "area under the serum
concentration curve from time zero to the time of the **last quantifiable**
concentration".

```{r pknca}
# Filter to !is.na() only; never add `time > 0` or `Cc > 0` -- both drop the
# time-zero anchor PKNCA needs for AUC.
sim_nca <- strata_sim |>
  filter(!is.na(Cc)) |>
  group_by(id, treatment) |>
  filter(time <= max(time[Cc >= lloq_ug_ml])) |>   # to the last quantifiable sample
  ungroup() |>
  select(id, time, Cc, treatment)

# Guarantee a time = 0 row per subject; pre-dose Cc = 0 is correct for SC dosing.
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(sim_nca), Cc ~ time | treatment + id)

dose_df <- strata_sim |>
  distinct(id, treatment) |>
  tidyr::expand_grid(time = c(0, 14, 28)) |>
  mutate(amt = 140) |>
  as.data.frame()
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(start = 0, end = Inf,
                        cmax = TRUE, tmax = TRUE, auclast = TRUE)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

## Comparison against published NCA

```{r nca-compare}
published <- tibble::tribble(
  ~treatment,                       ~cmax, ~auclast,
  "Low-moderate intensity statin",   20,     226,
  "High intensity statin",           16,     181
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by        = "treatment",
  units     = c(cmax = "ug/mL", auclast = "day*ug/mL", tmax = "day"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste("Simulated typical-value versus published NCA for 140 mg SC Q2W",
                  "(Gibbs 2017 Table 1). * differs from reference by >20%.")
)
```

**Cmax agrees** with the published means in both strata (within about 12%).
**AUClast is starred**, and the reason is structural rather than a
transcription error -- it is investigated, not tuned, below.

```{r nca-ratio}
nca_wide <- as.data.frame(nca_res$result) |>
  filter(PPTESTCD %in% c("cmax", "auclast")) |>
  select(treatment, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

ratios <- tibble::tibble(
  metric     = c("Cmax", "AUClast"),
  simulated  = c(
    nca_wide$cmax[nca_wide$treatment == "High intensity statin"] /
      nca_wide$cmax[nca_wide$treatment == "Low-moderate intensity statin"],
    nca_wide$auclast[nca_wide$treatment == "High intensity statin"] /
      nca_wide$auclast[nca_wide$treatment == "Low-moderate intensity statin"]
  ),
  published  = c(0.80, 0.80)
)

ratios |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  dplyr::rename("Metric" = metric,
                "Simulated ratio (high / low-moderate)" = simulated,
                "Published ratio" = published) |>
  knitr::kable(caption = paste("Target-mediated-elimination contrast: the effect of a 26%",
                               "higher baseline PCSK9 on unbound evolocumab exposure."))

stopifnot(
  # Deterministic typical-value profiles, so a tight bound is appropriate.
  # This is the sharpest available check on the target-mediated elimination
  # term: it depends on kint, kss, BASE_PCSK9 and V jointly, and is independent
  # of the sampling schedule and of the mg -> nmol conversion constant.
  all(abs(ratios$simulated - ratios$published) < 0.10),
  # Cmax itself, against the published means.
  abs(nca_wide$cmax[nca_wide$treatment == "Low-moderate intensity statin"] - 20) / 20 < 0.25,
  abs(nca_wide$cmax[nca_wide$treatment == "High intensity statin"]         - 16) / 16 < 0.25
)
```

### Why AUClast is high and Cmax is not

The comparator in Table 1 is an **arithmetic mean over 6 and 9 real subjects**,
whereas the simulation above is the **typical (median) subject**. With
`omega^2(CL) = 0.924` these are not the same quantity: for a log-normal
parameter the mean is `exp(omega^2 / 2)` times the median, so the mean
subject's clearance is about 1.6-fold the typical subject's, and the same
argument applies to `kss` (`omega^2 = 1.17`).

```{r skew}
c(CL_mean_over_median  = exp(0.924 / 2),
  kss_mean_over_median = exp(1.170 / 2))
```

AUC is inversely proportional to clearance and so absorbs this skew almost in
full, while Cmax -- set largely by dose, volume and absorption -- is far less
sensitive to it. Two further effects push the same way and only on AUClast:
each real subject's `auclast` stops at *their own* last quantifiable sample, so
faster-clearing subjects contribute truncated areas, and the published
dispersion is enormous (226 +/- 249 with n = 6; the SD exceeds the mean), so the
observed mean is itself imprecisely estimated. The ratio check above is
immune to all three, which is why it is the assertion that is gated.

# Assumptions and deviations

* **Evolocumab molecular weight (141,800 g/mol) is not reported in Gibbs
  2017.** The paper works entirely in molar units and never prints a
  mg <-> nmol bridge, but doses are administered in mg, so one is required to
  make the model dosable in the paper's own units. The value used is from the
  FDA-approved Repatha (evolocumab) prescribing information, and is the same
  constant already carried by the sibling registry model
  `Kuchimanchi_2018_evolocumab.R`. It is annotated as non-paper provenance at
  its definition in `model()`. Its influence on the paper's reported endpoints
  is small: raising it to 150,000 g/mol (a 5.8% change) moves the three
  time-averaged LDL-C reductions by less than 0.9 percentage points, and it
  cancels entirely out of the Cmax and AUClast **ratios** gated above.
* **PCSK9 molecular weight (about 71.9 kDa)** is *not* an external assumption:
  it is recovered from the paper's own two ng/mL <-> nM statements (379 ng/mL =
  5.27 nM and 242 ng/mL = 3.36 nM), which agree to within 0.15%. It is used
  only to convert Table 1's baseline PCSK9 values into the model's molar scale.
* **Equation 5 was re-derived.** The PDF text layer drops the radical and the
  squared term from the printed equation 5. The expression used here is the
  positive root of the QSS binding quadratic, which reproduces the printed
  `4 kss TDC` term exactly and returns `FDC = TDC` in the no-target limit and
  `FDC = 0` at zero dose (both verified numerically above).
* **`k` in equation 2 is read as `kel = CL/V`.** The paper writes the linear
  elimination term as `k * FDC * V` while Table 2 reports `CL`, and the Figure 1
  legend defines `kel` as the elimination rate constant; `k * FDC * V` is
  therefore `CL * FDC`, which is how it is encoded.
* **Table S1 is not on disk.** The paper's baseline-demographics table is in
  the Supporting Information, which is not part of the open-access deposit. The
  population metadata therefore uses the pooled demographic summaries stated in
  the Results narrative (n, sex, race, mean age, mean weight) rather than
  per-study strata. No model parameter depends on Table S1.
* **The statin-intensity strata are simulated by setting `lrbase_target`
  directly**, not through `DIS_HEALTHY`. Statin *intensity* is not a covariate
  in the final model -- Gibbs 2017 reports the Table 1 contrast as observed NCA
  in two subgroups that differ in measured baseline PCSK9 -- so the honest
  reproduction sets each subgroup's observed baseline PCSK9 and lets the TMDD
  structure generate the exposure difference.
* **Two single-dose PCSK9-suppression details are known deviations** and are
  excluded from the assertion gate rather than having the bounds widened around
  them. The typical-value prediction gives 76% maximum suppression after a
  single 21 mg SC dose where the paper reports "> 80%", and holds full
  suppression (below the 15 ng/mL assay LLOQ) for about 21 days after 420 mg
  where the paper reports a maximum of 14 days. Both published figures describe
  *observed* individual data sampled on a discrete visit schedule whose
  resolution near two weeks is about a week (days 8, 11, 15, 22, 29), and
  observed suppression is left-censored at the assay LLOQ, so neither is
  directly comparable to a continuous typical-value trajectory. Nothing in the
  model was tuned toward either number; the dose-ordering, the depth of
  suppression at 70 mg and above, and the absence of suppression at 7 mg all
  reproduce.
* **The published AUClast comparison is a known, explained deviation** and is
  deliberately excluded from the assertion gate; see the section above. The
  schedule- and unit-independent Cmax and AUClast *ratios* are gated instead.
* **No covariates other than `DIS_HEALTHY` are implemented**, because the final
  model retains none. Body weight in particular was not retained, so no
  allometric scaling is applied.
* **Between-subject variability is not applied to `kdeg`, `kint`, `theta 1`,
  `kout`, or `Imax`**, matching the paper: Tables 2-3 report 0.00 with the
  footnote "Intersubject random variance was fixed at 0", and Methods states
  "BSV for kdeg and kint were fixed at 0%".
* **The model has no IIV correlation between the PK and PD layers.** The paper
  reports a covariance block only for ka/V/CL; `BASE_PCSK9`, `kss`,
  `BASE_LDL-C` and `IC50` are given as diagonal variances and are encoded that
  way.
