---
title: "Pemigatinib population PK and exposure-response (Gong 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pemigatinib population PK and exposure-response (Gong 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Gong 2023 reports two things: a refined population PK model for pemigatinib, and a
set of exposure-response analyses built on top of it. Two of those models are
packaged here.

* Citation: `r rxode2::rxode(readModelDb("Gong_2023_pemigatinib"))$reference`
* Article: <https://doi.org/10.1002/psp4.13064>
* Supplement (Appendix S1, containing Table S1, Table S2 and the final NONMEM
  control stream): <https://doi.org/10.1002/psp4.13064>

| Model | Role |
|---|---|
| `Gong_2023_pemigatinib` | Two-compartment population PK with sequential zero-/first-order absorption |
| `Gong_2023_pemigatinib_creatinine` | Direct-effect Emax model for percentage change from baseline in serum creatinine, driven by steady-state AUC |

```{r models}
mod  <- readModelDb("Gong_2023_pemigatinib")
modc <- readModelDb("Gong_2023_pemigatinib_creatinine")
```

Pemigatinib is a selective inhibitor of fibroblast growth factor receptor (FGFR)
1-3, approved for previously treated, unresectable locally advanced or metastatic
cholangiocarcinoma (CCA) with *FGFR2* fusions or rearrangements, and for
relapsed/refractory myeloid/lymphoid neoplasms with *FGFR1* rearrangement. The
recommended regimen is 13.5 mg orally once daily on a 21-day cycle, 2 weeks on and
1 week off.

# Population

The population PK analysis pooled 4552 concentration records from 467 healthy
participants and patients across seven FIGHT studies (FIGHT-101, -102, -104, -105,
-106, -202, -203; Gong 2023 Table S1). Mean (SD) age was 54.3 (14.5) years with a
median of 56 and range 19-83; mean (SD) baseline body weight was 75.7 (19.7) kg
with a median of 73.9 kg and range 39.8-156 kg; 219 of 467 (46.9%) were men, so
women are the more common -- and therefore reference -- sex category. Tumour types
were CCA (163; 34.9%), myeloid/lymphoid neoplasms with *FGFR1* rearrangement (34;
7.3%), other cancers (192; 41.1%), and healthy participants (78; 16.7%). Relevant
co-medication use was phosphate binders in 61 (13.1%), proton-pump inhibitors
(PPI) in 114 (24.4%), and histamine-2 receptor antagonists (H2RA) in 39 (8.4%).
89.3% received pemigatinib monotherapy. Baseline demographics are Gong 2023
Table 1.

The exposure-response analysis of serum creatinine used the 300 monotherapy
patients pooled from FIGHT-101, FIGHT-102, and FIGHT-202.

The same information is available programmatically via
`readModelDb("Gong_2023_pemigatinib")()$population`.

# Source trace

Per-parameter origins are recorded as in-file comments next to each `ini()` entry
in `inst/modeldb/specificDrugs/Gong_2023_pemigatinib.R` and
`inst/modeldb/specificDrugs/Gong_2023_pemigatinib_creatinine.R`. They are
collected here for review.

## Population PK model

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` (CL/F) | 10.7 L/h | Table 2, %RSE 2.7, 95% CI 10.1-11.3 |
| `lvc` (Vc/F) | 118 L | Table 2, %RSE 3.4, 95% CI 110-125 |
| `lvp` (Vp/F) | 95.0 L | Table 2, %RSE 3.0, 95% CI 89.4-101 |
| `lq` (Q/F) | 25.2 L/h | Table 2, %RSE 7.7, 95% CI 21.4-29.0 |
| `lka` (ka) | 3.67 1/h | Table 2, %RSE 12.2, 95% CI 2.79-4.55 |
| `ld1` (D1) | 0.810 h | Table 2, %RSE 3.0, 95% CI 0.763-0.857 |
| `e_phosbinder_cl` | -0.155 | Table 2 "Phosphate binder on CL/F, %" = -15.5% |
| `e_male_cl` | +0.262 | Table 2 "Male sex on CL/F, %" = +26.2% |
| `e_ppi_ka` | -0.620 | Table 2 "PPI on Ka, %" = -62.0% |
| `e_male_ka` | -0.583 | Table 2 "Male sex on Ka, %" = -58.3% |
| `e_h2ra_d1` | -0.334 | Table 2 "H2B antagonist on D1, %" = -33.4% |
| `e_wt_vc` | 0.842 | Table 2; Appendix S1 `V2BWT = ((BWT/73.9)**THETA(12))` |
| `e_wt_vp` | 1.13 | Table 2; Appendix S1 `V3BWT = ((BWT/73.9)**THETA(13))` |
| Reference weight | 73.9 kg | Appendix S1 control stream; = Table 1 cohort median |
| `etalcl`, `etalvc` block | 0.200704, 0.129, 0.190969 | Table 2 IIV 44.8 / 43.7 %CV; covariance 0.129 from Table 2 footnote a |
| `etalka` | 1.69 | Table 2 IIV Ka 130 %CV |
| `etald1` | 0.370881 | Table 2 IIV D1 60.9 %CV |
| IIV on Q/F, Vp/F | fixed to 0 | Table 2 "Fixed to 0"; Appendix S1 `$OMEGA 0 FIX` |
| `propSd` | 0.323 | Table 2 RUV 32.3%; Appendix S1 `$ERROR` log-additive |
| Absorption structure | sequential zero-order D1 then first-order ka | Appendix S1 `$SUBROUTINE ADVAN4 TRANS4` with `D1=TVD1*EXP(ETA(6))`; Results paragraph 2 |
| `Cc` unit conversion | `x 1e6 / 487.5` | Appendix S1 `S2 = V2/1000000*487.5 ;nM` (pemigatinib MW 487.5 g/mol) |

## Serum-creatinine exposure-response model

| Equation / parameter | Value | Source location |
|---|---|---|
| `lemax` | 43.4 % | Table S2, approximate SE 6.15, 95% CL 31.3-55.5 |
| `lec50` | 2595 h*nM | Table S2, approximate SE 793, 95% CL 1035-4154 |
| `creatPctChange = Emax * AUC / (EC50 + AUC)` | n/a | Methods "Exposure-response evaluation" |
| Residual error | not reported | absent from Table S2 and main text; encoded `fixed(0)` |

## Interpreting the reported %CV as a log-scale SD

Gong 2023 Table 2 reports inter-individual variability as "%CV" without saying
whether those numbers are `100 * omega` (the log-scale SD) or
`100 * sqrt(exp(omega^2) - 1)` (the log-normal coefficient of variation). The
paper's own numbers settle it: Table 2 gives the CL/F-Vc/F correlation as 0.659,
and the footnote gives the covariance as 0.129. Only one reading reproduces that
correlation.

```{r iiv-scale}
cv_cl <- 0.448  # Table 2 IIV CL/F 44.8 %CV
cv_vc <- 0.437  # Table 2 IIV Vc/F 43.7 %CV
cov_cl_vc <- 0.129  # Table 2 footnote a

tibble::tibble(
  Reading = c("omega^2 = CV^2 (log-scale SD)", "omega^2 = log(1 + CV^2) (log-normal CV)"),
  `Implied correlation` = c(
    cov_cl_vc / (cv_cl * cv_vc),
    cov_cl_vc / sqrt(log(1 + cv_cl^2) * log(1 + cv_vc^2))
  ),
  `Reported (Table 2)` = 0.659
) |>
  knitr::kable(digits = 4, caption = "The reported %CV values are log-scale SDs.")
```

The Appendix S1 `$OMEGA` initial estimates corroborate the same reading: the ka
initial 1.65573 has square root 1.287 (128.7%) against the 130% final, and the D1
initial 0.429359 has square root 0.655 (65.5%) against the 60.9% final. The model
file therefore encodes `omega^2 = CV^2`.

```{r iiv-scale-check}
stopifnot(abs(cov_cl_vc / (cv_cl * cv_vc) - 0.659) < 0.002)
```

# Virtual cohort

Original observed data are not publicly available. The stochastic simulation
below uses a virtual population whose body-weight distribution approximates the
published trial demographics (Gong 2023 Table 1): drawn log-normally with a
median of 73.9 kg and truncated to the observed 39.8-156 kg range. All other
covariates are set to the reference level (female, no concomitant
phosphate binder, proton-pump inhibitor, or H2-receptor antagonist).

Covariate effects are examined separately, in the *typical-value* section further
down, because Gong 2023 states its own co-medication results as typical-subject
simulations. A deterministic comparison isolates the published coefficients
exactly; a stochastic one would only add sampling noise, and with 130% CV on `ka`
that noise is large.

```{r cohort}
set.seed(20231101)

n_subj <- 200L
dose_mg <- 13.5
n_days <- 8L  # terminal t1/2 is ~15 h, so 8 days is >12 half-lives to steady state
tau <- 24
ss_start <- (n_days - 1) * tau

subj <- tibble::tibble(
  id = seq_len(n_subj),
  WT = pmin(pmax(rlnorm(n_subj, meanlog = log(73.9), sdlog = 0.25), 39.8), 156),
  SEXF = 1,
  CONMED_PPI = 0,
  CONMED_H2RA = 0,
  CONMED_PHOSBINDER = 0
)

doses <- subj |>
  tidyr::crossing(time = seq(0, by = tau, length.out = n_days)) |>
  dplyr::mutate(
    amt = dose_mg,
    evid = 1L,
    # rate = -2 tells rxode2 to apply the model's dur(depot) zero-order window
    # rather than treating the dose as an instantaneous bolus.
    rate = -2,
    cmt = "depot"
  )

obs <- subj |>
  tidyr::crossing(time = ss_start + seq(0, tau, by = 0.25)) |>
  dplyr::mutate(
    amt = NA_real_,
    evid = 0L,
    rate = NA_real_,
    # Observations sit on the ODE state, never on the algebraic observable Cc
    # (which would auto-inject a cmt slot and renumber the states).
    cmt = "central"
  )

events <- dplyr::bind_rows(doses, obs) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(mod, events = events, keep = c("WT")) |>
  as.data.frame()
```

# Replicate published figures

## Steady-state concentration-time profile (structure of Figure 1)

Gong 2023 Figure 1 is a prediction-corrected and an ordinary visual predictive
check over the whole pooled dataset. The published figure cannot be reproduced
without the observed data, but the simulated median and 5th/95th percentiles over
a steady-state interval at the recommended 13.5 mg once-daily dose show the
behaviour the model encodes: a zero-order release window of `D1` followed by
rapid first-order absorption, and a biphasic decline.

```{r figure-1}
sim |>
  dplyr::mutate(tad = time - ss_start) |>
  dplyr::group_by(tad) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  labs(
    x = "Time after dose (h)", y = "Pemigatinib concentration (nM)",
    title = "Simulated steady-state profile, 13.5 mg once daily",
    caption = "Structure underlying Figure 1 of Gong 2023 (median with 5th-95th percentiles, n = 200)."
  )
```

## Serum creatinine vs steady-state exposure (Figure 4)

Gong 2023 Figure 4 plots model-predicted percentage change from baseline in serum
creatinine against steady-state AUC. The paper states the predicted values at four
doses (Results, "Exposure-response analysis of safety"), obtained from the mean
post hoc AUCss at 13.5 mg once daily and the dose-scaled means at the other doses.

```{r figure-4}
auc_ref_135 <- 2850  # Gong 2023 Results: mean AUCss at 13.5 mg q.d.

# A plotting grid for the algebraic exposure-response curve -- these are
# evaluation points, not virtual participants.
creat_curve <- tibble::tibble(
  id = seq_len(200),
  time = 0,
  AUC_PEMI = seq(0, 9000, length.out = 200)
)

creat_sim <- rxode2::rxSolve(modc, events = creat_curve, keep = "AUC_PEMI") |>
  as.data.frame()

dose_points <- tibble::tibble(
  id = 1:4,
  time = 0,
  dose_mg = c(6, 9, 13.5, 20),
  AUC_PEMI = auc_ref_135 * c(6, 9, 13.5, 20) / 13.5
)

dose_pred <- rxode2::rxSolve(modc, events = dose_points, keep = c("AUC_PEMI", "dose_mg")) |>
  as.data.frame()

ggplot(creat_sim, aes(AUC_PEMI, creatPctChange)) +
  geom_line(linewidth = 0.8) +
  geom_point(data = dose_pred, aes(AUC_PEMI, creatPctChange),
             size = 2.5, colour = "firebrick") +
  labs(
    x = "Steady-state AUC (h*nM)",
    y = "Change from baseline in serum creatinine (%)",
    title = "Serum creatinine exposure-response",
    caption = "Replicates Figure 4 of Gong 2023; points are the 6, 9, 13.5 and 20 mg doses."
  )
```

The four published predictions are reproduced exactly.

```{r figure-4-check}
creat_cmp <- dose_pred |>
  dplyr::transmute(
    `Dose (mg)` = dose_mg,
    `AUCss (h*nM)` = round(AUC_PEMI, 1),
    Simulated = round(creatPctChange, 2),
    Published = c(14.2, 18.3, 22.7, 26.9)
  ) |>
  dplyr::mutate(`Difference (%)` = round(100 * (Simulated - Published) / Published, 2))

knitr::kable(
  creat_cmp,
  caption = "Serum creatinine percentage change: packaged model vs Gong 2023 Results."
)

# Strict gate: every published prediction matched to within rounding.
stopifnot(all(abs(creat_cmp$Simulated - creat_cmp$Published) <= 0.05))
```

# PKNCA validation

Noncompartmental parameters are computed over the final (steady-state) dosing
interval.

```{r pknca}
sim_nca <- sim |>
  # Only !is.na(Cc): adding time > 0 or Cc > 0 would drop the anchor row PKNCA
  # needs at the interval start.
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc) |>
  dplyr::mutate(regimen = "13.5 mg q.d.")

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | regimen + id)

dose_df <- events |>
  dplyr::filter(evid == 1, time == ss_start) |>
  dplyr::select(id, time, amt) |>
  dplyr::mutate(regimen = "13.5 mg q.d.")

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | regimen + id)

intervals <- data.frame(
  start = ss_start,
  end = ss_start + tau,
  cmax = TRUE,
  tmax = TRUE,
  cmin = TRUE,
  auclast = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_tab <- as.data.frame(nca_res) |>
  # PKNCA emits dependency rows; restrict to the requested interval.
  dplyr::filter(start == ss_start, end == ss_start + tau) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "cmin", "auclast")) |>
  dplyr::select(regimen, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

nca_tab |>
  dplyr::summarise(
    `Cmax,ss (nM)` = mean(cmax),
    `Cmin,ss (nM)` = mean(cmin),
    `AUCss (h*nM)` = mean(auclast),
    `Median Tmax (h)` = median(tmax)
  ) |>
  knitr::kable(digits = 1,
               caption = "Simulated steady-state exposure, 13.5 mg once daily (n = 200).")
```

## Mean steady-state AUC at the recommended dose

Gong 2023 reports a mean AUCss of 2850 h*nM at 13.5 mg once daily, and separately
notes that the serum-phosphate EC50 of 1665 h*nM "approximates the mean AUC0-24h
from 8 mg q.d. pemigatinib". Both are reproduced.

```{r auc-check}
auc_mean <- mean(nca_tab$auclast)

auc_cmp <- tibble::tibble(
  Quantity = c("Mean AUCss at 13.5 mg", "Implied mean AUCss at 8 mg"),
  Simulated = c(auc_mean, auc_mean * 8 / 13.5),
  Published = c(2850, 1665),
  Source = c("Results, exposure-response of efficacy",
             "Results: phosphate EC50 1665 h*nM ~ 8 mg q.d.")
) |>
  dplyr::mutate(`Difference (%)` = 100 * (Simulated - Published) / Published)

knitr::kable(auc_cmp, digits = c(0, 0, 0, 0, 1),
             caption = "Simulated vs published steady-state exposure anchors.")

stopifnot(all(abs(auc_cmp$`Difference (%)`) < 10))
```

The arithmetic mean AUCss exceeds the typical-value AUCss because clearance is
log-normally distributed: `E[1/CL] = (1/CL_typical) * exp(omega^2 / 2)`, which for
`omega^2 = 0.2007` inflates the mean by 10.6%.

## Dose conservation

For the typical reference subject, AUCss over the dosing interval must equal
`dose / (CL/F)` converted to nM. This checks that the sequential zero-/first-order
absorption model loses no dose.

```{r mass-balance}
mod_typ <- rxode2::zeroRe(mod)

typ_events <- events |>
  dplyr::filter(id == 1L) |>
  dplyr::mutate(WT = 73.9)

sim_typ <- rxode2::rxSolve(mod_typ, events = typ_events) |>
  as.data.frame() |>
  dplyr::filter(time >= ss_start)

auc_trapz <- sum(diff(sim_typ$time) *
                   (head(sim_typ$Cc, -1) + tail(sim_typ$Cc, -1)) / 2)
auc_closed <- dose_mg / 10.7 * 1e6 / 487.5

tibble::tibble(
  `Simulated AUCss (h*nM)` = auc_trapz,
  `Closed form dose/(CL/F) (h*nM)` = auc_closed,
  `Difference (%)` = 100 * (auc_trapz - auc_closed) / auc_closed
) |>
  knitr::kable(digits = c(1, 1, 4),
               caption = "Dose conservation through the sequential zero-/first-order absorption model.")

stopifnot(abs(auc_trapz - auc_closed) / auc_closed < 0.005)
```

# Typical-subject covariate effects

Gong 2023 Results states: "Simulated Cmax,ss values were 12% lower and 1.1%
higher, respectively, in **typical** PPI or H2B users versus nonusers." Because
those are typical-value statements, they are reproduced deterministically with the
random effects zeroed, at the cohort-median weight of 73.9 kg.

```{r typical-covariates}
typ_profile <- function(sexf, ppi, h2ra, binder) {
  # A single deterministic subject, so a dense grid is cheap. The 0.25 h grid
  # used for the stochastic cohort is too coarse to resolve Cmax to the one
  # decimal place the paper reports.
  ev_dose <- events |> dplyr::filter(id == 1L, evid == 1L)
  ev_obs <- ev_dose[1, ] |>
    dplyr::select(id, amt, evid, rate, cmt) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, rate = NA_real_, cmt = "central") |>
    tidyr::crossing(time = ss_start + seq(0, tau, by = 0.01))

  ev <- dplyr::bind_rows(ev_dose, ev_obs) |>
    dplyr::arrange(time, dplyr::desc(evid)) |>
    dplyr::mutate(
      WT = 73.9,
      # Lower-case argument names so the dplyr data mask cannot shadow them
      # with the same-named event-table columns.
      SEXF = sexf,
      CONMED_PPI = ppi,
      CONMED_H2RA = h2ra,
      CONMED_PHOSBINDER = binder
    )
  s <- rxode2::rxSolve(mod_typ, events = ev) |>
    as.data.frame() |>
    dplyr::filter(time >= ss_start)
  c(
    cmax = max(s$Cc),
    cmin = min(s$Cc),
    auc = sum(diff(s$time) * (head(s$Cc, -1) + tail(s$Cc, -1)) / 2)
  )
}

ref_t    <- typ_profile(1, 0, 0, 0)
male_t   <- typ_profile(0, 0, 0, 0)
ppi_t    <- typ_profile(1, 1, 0, 0)
h2ra_t   <- typ_profile(1, 0, 1, 0)
binder_t <- typ_profile(1, 0, 0, 1)

pct <- function(a, b, m) 100 * (a[m] / b[m] - 1)

typ_cmp <- tibble::tibble(
  Comparison = c("PPI user vs nonuser", "H2RA user vs nonuser",
                 "Phosphate-binder user vs nonuser", "Women vs men"),
  `Cmax,ss (%)` = c(pct(ppi_t, ref_t, "cmax"), pct(h2ra_t, ref_t, "cmax"),
                    pct(binder_t, ref_t, "cmax"), pct(ref_t, male_t, "cmax")),
  `Cmin,ss (%)` = c(pct(ppi_t, ref_t, "cmin"), pct(h2ra_t, ref_t, "cmin"),
                    pct(binder_t, ref_t, "cmin"), pct(ref_t, male_t, "cmin")),
  `AUCss (%)` = c(pct(ppi_t, ref_t, "auc"), pct(h2ra_t, ref_t, "auc"),
                  pct(binder_t, ref_t, "auc"), pct(ref_t, male_t, "auc")),
  `Published Cmax,ss (%)` = c(-12, 1.1, 6, 35)
)

knitr::kable(typ_cmp, digits = 1,
             caption = "Typical-subject steady-state exposure changes (Gong 2023 Results).")
```

The two statements the paper makes explicitly about a **typical** subject are
matched to the reported precision.

```{r typical-covariates-check}
stopifnot(abs(pct(ppi_t,  ref_t, "cmax") - (-12)) < 0.5)
stopifnot(abs(pct(h2ra_t, ref_t, "cmax") - 1.1)   < 0.5)
# Sex and phosphate-binder effects on AUC follow directly from the CL
# coefficients: +26.2% for women vs men, and 1/(1 - 0.155) - 1 = +18.3% for
# phosphate-binder users.
stopifnot(abs(pct(ref_t, male_t, "auc") - 26.2) < 0.3)
stopifnot(abs(pct(binder_t, ref_t, "auc") - (1 / (1 - 0.155) - 1) * 100) < 0.3)
```

The sex comparison needs a caveat. The AUC ratio matches the published CL
coefficient exactly (+26.2% in women), and the phosphate-binder Cmax change
(+7.8% typical) brackets the paper's population estimate of "approximately 6%".
But the paper reports Cmax,ss +35% and Cmin,ss +16% in women versus men, whereas
this typical-value comparison -- which holds body weight fixed at 73.9 kg for both
sexes -- gives +24.5% and +47.1%. The paper's own Discussion supplies the
explanation: the higher exposures in women were "possibly due to men having higher
body weight than women and consequently higher Vc/F". Gong 2023 Table 1 reports
weight only for the pooled cohort, not by sex, so the sex-specific weight
distributions needed to reproduce the confounded comparison are not available.
Holding weight equal isolates the published sex coefficients themselves, which are
reproduced exactly on AUC. No parameter was tuned to close the gap.

# Assumptions and deviations

* **Reported "%CV" is a log-scale SD.** Gong 2023 Table 2 labels the IIV column
  "%CV" without defining the scale. The model encodes `omega^2 = CV^2` because
  that reading -- and only that reading -- reproduces the paper's own reported
  CL/F-Vc/F correlation of 0.659 from the reported covariance of 0.129 (see
  "Interpreting the reported %CV" above). The Appendix S1 `$OMEGA` initial
  estimates independently corroborate it.
* **The two body-weight rows in Table 2 are exponents, not percentages.** Table 2
  labels the rows "Body weight on Vc/F, %" and "Body weight to Vp/F, %", but the
  Appendix S1 control stream writes them as `((BWT/73.9)**THETA)`, i.e.
  dimensionless power exponents. The "%" is a formatting artifact inherited from
  the categorical covariate rows above them in the same column.
* **Sex coding is inverted relative to the source.** The source column `SEXN` is
  1 = male, 2 = female, with female the reference ("Most common") level. The
  canonical covariate is `SEXF` (1 = female), so the model applies the male
  deviation via the `(1 - SEXF)` indicator. Female remains the reference.
* **Reference weight 73.9 kg** is the cohort median from Table 1, taken from the
  control stream rather than a rounded standard weight.
* **Sex comparison holds body weight equal.** See the caveat above: the paper's
  reported Cmax,ss / Cmin,ss sex differences are confounded with body weight,
  which Table 1 does not report by sex.
* **Residual variability for the exposure-response model is not reported.**
  Gong 2023 Figure 4 shows simulated 5th and 95th percentiles, so the fitted
  creatinine model carried a residual term, but neither Table S2 nor the main
  text gives its magnitude. It is encoded as `fixed(0)` rather than invented, so
  `Gong_2023_pemigatinib_creatinine` is a typical-value (deterministic)
  predictor.
* **No inter-individual variability is reported for the exposure-response model**
  either; it is a population-level regression on per-subject AUCss.
* **Known convention deviation: the exposure-response output name.**
  `checkModelConventions()` emits one warning for
  `Gong_2023_pemigatinib_creatinine` -- that the single-output observation
  variable `creatPctChange` is neither `Cc` nor a registered PD-output canonical.
  The endpoint is a percentage change from baseline in a serum chemistry analyte:
  not a drug concentration, and not backed by an ODE state, so no existing
  canonical compartment name fits. The model declares
  `paper_specific_compartments <- c("creatPctChange")`, which suppresses the
  compartment warning but not the observation warning. Registering a canonical
  name for change-from-baseline serum-chemistry endpoints is pending operator
  ratification; the name may change when that is settled. The population PK model
  `Gong_2023_pemigatinib` lints clean.
* **`AUC_PEMI` is a static covariate, not an integrated state.** This mirrors the
  authors' own workflow: AUCss was derived noncompartmentally from a dense
  simulated steady-state profile in NONMEM and then used as a regressor in SAS.
  The vignette chains the two layers by taking AUCss from the packaged population
  PK model and feeding it to the creatinine model.
* **The virtual cohort is not the observed cohort.** Weight is drawn log-normally
  around the published median and truncated to the published range; the observed
  joint distribution of weight, sex, and co-medication use is not published.
* **Dose-scaling of AUCss across dose groups** in the Figure 4 replication follows
  the paper, which states the predictions at 6, 9, and 20 mg were computed from
  "the dose-scaled mean post hoc AUCss" anchored on the 13.5 mg mean of
  2850 h*nM. Pemigatinib PK is linear in this model, so this is exact.

# Errata and reporting gaps

No erratum or corrigendum for Gong 2023 was located.

Several exposure-response sub-models reported in the paper are **not** packaged:

* **Serum phosphate Emax model.** Table S2 reports Emax = 5.51 mg/dL,
  EC50 = 1665 h*nM, and a baseline-serum-phosphate coefficient of -0.185 per
  mg/dL -- a complete parameter set. This model is deferred only pending
  ratification of a canonical covariate column for baseline serum phosphate: no
  serum-phosphate entry exists in `inst/references/covariate-columns.md`, and the
  choice between the paper's mg/dL and SI mmol/L needs an operator decision,
  since serum albumin was made SI-canonical. It is otherwise ready to package.
* **Objective-response-rate logistic regressions** (Figure 3a on serum-phosphate
  change, Figure 3b on AUCss). The paper reports only model-predicted ORR at four
  doses and describes the shape as bell-shaped; no intercept or coefficient is
  given, so the model cannot be reconstructed without fitting to the authors' own
  predictions.
* **Hyperphosphatemia logistic regression** (Figure 5). The log-AUCss coefficient
  (1.40) and the odds ratios for baseline phosphate (2.46), albumin (4.94), and
  *FGFR2* status (0.498) are reported, but the intercept and the covariate
  centering values are not, so absolute probabilities cannot be computed.
* **Cox proportional-hazards model for progression-free survival.** Reported as
  not statistically significant, with no parameter estimates.
