---
title: "Morphine brain distribution in the rat (Groenendaal 2007)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Morphine brain distribution in the rat (Groenendaal 2007)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r model}
mod <- rxode2::rxode(readModelDb("Groenendaal_2007_morphine_brain_rat"))
```

# Model and source

* Citation: `r mod$reference`
* Article: <https://doi.org/10.1038/sj.bjp.0707257>
* PubMed Central: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2014001/>
* Companion paper (separate model, not extracted here): Groenendaal *et al.*
  (2007) *Br J Pharmacol* **151**(4):713-720,
  <https://doi.org/10.1038/sj.bjp.0707258> -- the EEG PK-PD biophase model
  fitted to the same experiments.

`r mod$description`

# Population

The model was built from a single rat study at Leiden University
(paper Table 1). Seventy-one male Wistar rats (Charles River, Maastricht),
250-350 g, received a 10 min zero-order intravenous infusion of morphine
hydrochloride at 4, 10 or 40 mg/kg. The 4 mg/kg level was studied both with
vehicle and with a co-infusion of the P-glycoprotein inhibitor GF120918
(elacridar); 10 and 40 mg/kg were vehicle only. All animals received a
continuous midazolam infusion (5.5 mg/kg/h) to suppress opioid-induced seizure
activity, and animals in the 40 mg/kg arm were artificially ventilated with
vecuronium-bromide muscle relaxation because of severe respiratory depression.

The two model layers were fitted sequentially in NONMEM V (FOCE INTERACTION).
The blood three-compartment layer used arterial blood morphine from all 71
animals (15 samples per rat to 360 min). The brain layer used
recovery-corrected striatal microdialysate from the 32 animals perfused with
blank perfusate -- the 50 and 500 ng/mL retrodialysis animals yielded no usable
post-dose ECF data and were excluded (paper *Microdialysis probe recovery*).

The same information is available programmatically:

```{r population}
str(mod$population)
```

# Model structure

The blood layer is an ordinary three-compartment disposition model
(NONMEM `ADVAN11 TRANS4`), with body weight entering CL and the first
peripheral volume V2 through the paper's **centred linear** covariate form
(equation 7),

$$P_i = \theta_1 \left( 1 + \theta_2 \left( \mathrm{BW}_i - \mathrm{median\ BW} \right) \right)$$

which is a linear slope in kg, *not* an allometric power.

The brain layer is the paper's contribution. Equation 9 writes the mass balance
on the amount of morphine in brain extracellular fluid as three explicit
transport terms,

$$\frac{\mathrm{d}A_{ecf}}{\mathrm{d}t}
  = Q_{diff}\left(C_{blood} - C_{ecf}\right)
  + \frac{N_{max} C_{blood}}{C_{50} + C_{blood}}
  - Q_{eff} C_{ecf}$$

and equations 10-11 divide through by the brain ECF distribution volume
$V_{ecf}$ and name the aggregated constants
$k_{diff} = Q_{diff}/V_{ecf}$, $N^{*}_{max} = N_{max}/V_{ecf}$ and
$k_{eff} = Q_{eff}/V_{ecf}$:

$$\frac{\mathrm{d}C_{ecf}}{\mathrm{d}t}
  = k_{diff}\left(C_{blood} - C_{ecf}\right)
  + \frac{N^{*}_{max} C_{blood}}{C_{50} + C_{blood}}
  - k_{eff} C_{ecf}$$

$V_{ecf}$ is never identified, so every published brain parameter is already
volume-aggregated. The packaged model therefore carries `brain_ecf` as a
**concentration** state in ng/mL rather than an amount -- encoding an amount
would require inventing $V_{ecf}$.

The mechanism the paper is after: with $C_{50} = 9.92$ ng/mL, the active influx
is essentially saturated at every blood concentration the study produced, so it
behaves as a constant input of $N^{*}_{max}$. As the dose rises, that fixed
contribution is diluted by the dose-proportional passive term, and the
dose-normalised brain exposure *falls*.

# Source trace

Every equation and every `ini()` value, with its location in the paper.

```{r source-trace, echo = FALSE}
src <- tibble::tribble(
  ~Item, ~Value, ~Source,
  "Blood disposition structure", "3-compartment, ADVAN11 TRANS4", "Methods, *PK analysis of blood profiles*",
  "Body-weight covariate form", "P = t1 * (1 + t2 * (BW - median BW))", "Equation 7",
  "GF120918 covariate form", "P = t3 * (1 - GF) + t4 * GF", "Equation 8",
  "Brain ECF mass balance", "dAecf/dt, three transport terms", "Equation 9",
  "Brain ECF concentration form", "dCecf/dt after dividing by Vecf", "Equation 10",
  "Parameter aggregation", "kdiff, N*max, keff = Q../Vecf", "Equation 11",
  "IIV form", "P_i = P_typ * exp(eta), eta ~ N(0, omega^2)", "Equations 3-4",
  "Residual error form", "Cobs = Cpred * (1 + eps), eps ~ N(0, sigma^2)", "Equations 5-6",
  "lcl = log(20.0)", "Cl Intercept 20.0 mL/min (CV 5.6%)", "Table 2",
  "e_wt_cl = 5.35", "Cl Slope factor 5.35 /kg (CV 25.2%)", "Table 2",
  "lvc = log(68.1)", "V1 68.1 mL (CV 16.7%)", "Table 2",
  "lq = log(15.5)", "Q2 15.5 mL/min (CV 11.3%)", "Table 2",
  "lvp = log(739)", "V2 Intercept 739 mL (CV 7.6%)", "Table 2",
  "e_wt_vp = 8.50", "V2 Slope factor 8.50 /kg (CV 17.1%)", "Table 2",
  "lq2 = log(17.8)", "Q3 17.8 mL/min (CV 18.4%)", "Table 2",
  "lvp2 = log(133)", "V3 133 mL (CV 15.9%)", "Table 2",
  "etalcl ~ 0.129", "omega^2 Cl 0.129 (CV 17.2%)", "Table 2",
  "etalvp ~ 0.099", "omega^2 V2 0.099 (CV 24.7%)", "Table 2",
  "propSd = sqrt(0.074)", "Proportional error 0.074 (CV 10.2%)", "Table 2",
  "lkdiff_bbb = log(0.0014)", "kdiff 0.0014 /min (CV 12.6%)", "Table 3",
  "lkefflux_bbb = log(0.0195)", "keff -GF120918 0.0195 /min (CV 12.2%)", "Table 3",
  "e_conmed_elacridar_kefflux_bbb", "keff +GF120918 0.0113 /min (CV 25.4%)", "Table 3",
  "lNstarMax = log(0.658)", "N*max 0.658 ng/mL/min (CV 26.1%)", "Table 3 (units per Abstract)",
  "lc50 = log(9.92)", "C50 9.92 ng/mL (CV 71.5%)", "Table 3",
  "etalkdiff_bbb / etalkefflux_bbb block", "0.238 / cov 0.059 / 0.080", "Table 3",
  "propSd_Cbrain_ecf = sqrt(0.094)", "Proportional error 0.094 (CV 21.4%)", "Table 3"
)
knitr::kable(src)
```

# Virtual cohort

Three arms matching the paper's brain-ECF design, 30 rats each. Body weights are
drawn uniformly over the paper's stated 250-350 g range (the paper reports only
the range and per-group means, not the individual weights).

```{r cohort}
rxode2::rxSetSeed(20070430)

REF_WT   <- 0.300      # kg; stands in for the paper's unprinted `median BW`
N_PER_ARM <- 30L       # well under the 200-per-arm cap

arms <- tibble::tribble(
  ~arm,                  ~dose_mgkg, ~gf,
  "4 mg/kg vehicle",              4,   0,
  "4 mg/kg + GF120918",           4,   1,
  "40 mg/kg vehicle",            40,   0
) |>
  mutate(arm = factor(arm, levels = arm))

subj <- arms |>
  tidyr::crossing(subj_no = seq_len(N_PER_ARM)) |>
  mutate(
    id = dplyr::row_number(),
    WT = runif(dplyr::n(), min = 0.250, max = 0.350),
    CONMED_ELACRIDAR = gf,
    amt = dose_mgkg * WT * 1e6      # ng
  )

# Observation grid: dense through the infusion and distribution phase, then
# every 15 min to 360 min. 165 min is included explicitly because the paper's
# dose-normalised AUC is reported over 0-165 min.
obs_times <- sort(unique(c(
  0, 1, 2, 3, 5, 7.5, 10, 12, 15, 20, 25, 30, 40, 50, 60,
  seq(75, 360, by = 15), 165
)))
```

The two observables are separate nlmixr2 endpoints (`Cc` is endpoint 1,
`Cbrain_ecf` endpoint 2), so every observation row carries a `dvid`. The row is
addressed at the real ODE state `central`; rxode2 maps `dvid` 1 and 2 onto the
endpoint pseudo-compartments it appended after the four ODE states, which the
solved output reports as `CMT` 5 and 6.

```{r events}
dose_rows <- subj |>
  transmute(id, arm, WT, CONMED_ELACRIDAR,
            time = 0, amt, rate = amt / 10,   # 10 min zero-order infusion
            evid = 1L, cmt = "central", dvid = NA_integer_)

obs_rows <- subj |>
  select(id, arm, WT, CONMED_ELACRIDAR) |>
  tidyr::crossing(time = obs_times, dvid = 1:2) |>
  mutate(amt = NA_real_, rate = NA_real_, evid = 0L, cmt = "central")

events <- bind_rows(dose_rows, obs_rows[names(dose_rows)]) |>
  arrange(id, time, dvid)
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod, events,
  keep = c("arm", "WT", "CONMED_ELACRIDAR"),
  useLinCmt = FALSE,           # ODE -> linCmt auto-conversion breaks dvid mapping
  returnType = "data.frame"
) |>
  as_tibble()

# CMT 5 rows are the Cc endpoint, CMT 6 rows the Cbrain_ecf endpoint; `sim`
# carries the residual error for whichever endpoint the row belongs to, while
# Cc / Cbrain_ecf are the individual predictions.
blood <- sim |>
  filter(CMT == min(CMT)) |>
  # `cl` is the solver's own INDIVIDUAL clearance for this subject (typical
  # value * exp(etalcl) * the body-weight factor); it is what the blood
  # identity check below must be compared against.
  transmute(id, arm, WT, time, cl, ipred = Cc, obs = sim)

ecf <- sim |>
  filter(CMT == max(CMT)) |>
  transmute(id, arm, WT, time, ipred = Cbrain_ecf, obs = sim)

nrow(blood)
nrow(ecf)
```

Typical-value (no IIV, no residual error) profiles at the reference weight, used
for every deterministic check below.

```{r typical}
typ_events <- arms |>
  mutate(id = dplyr::row_number(), WT = REF_WT,
         CONMED_ELACRIDAR = gf, amt = dose_mgkg * REF_WT * 1e6) |>
  select(id, arm, WT, CONMED_ELACRIDAR, amt, dose_mgkg)

typ_dose <- typ_events |>
  transmute(id, arm, WT, CONMED_ELACRIDAR, time = 0, amt, rate = amt / 10,
            evid = 1L, cmt = "central", dvid = NA_integer_)

typ_obs <- typ_events |>
  select(id, arm, WT, CONMED_ELACRIDAR) |>
  tidyr::crossing(time = sort(unique(c(seq(0, 360, by = 0.5), 165))), dvid = 1:2) |>
  mutate(amt = NA_real_, rate = NA_real_, evid = 0L, cmt = "central")

typ_sim <- rxode2::rxSolve(
  mod,
  bind_rows(typ_dose, typ_obs[names(typ_dose)]) |> arrange(id, time, dvid),
  keep = c("arm", "WT", "CONMED_ELACRIDAR"),
  omega = NA, sigma = NA, useLinCmt = FALSE, returnType = "data.frame"
) |>
  as_tibble() |>
  filter(CMT == min(CMT)) |>
  transmute(arm, time, Cc, Cbrain_ecf) |>
  left_join(select(typ_events, arm, dose_mgkg), by = "arm")
```

# Replicating the published figures

## Figure 2 -- morphine in blood

```{r fig2, fig.cap = "Replicates Figure 2 of Groenendaal 2007: simulated morphine blood concentrations by dose group. Points are simulated observations (individual prediction plus proportional residual error), the solid line the typical-value prediction, the ribbon the 2.5-97.5% simulated interval."}
blood_q <- blood |>
  group_by(arm, time) |>
  summarise(lo = quantile(obs, 0.025), hi = quantile(obs, 0.975), .groups = "drop")

ggplot(blood, aes(time, obs)) +
  geom_point(colour = "grey60", size = 0.5, alpha = 0.5) +
  geom_ribbon(data = blood_q, aes(time, ymin = pmax(lo, 1), ymax = hi),
              inherit.aes = FALSE, alpha = 0.2) +
  geom_line(data = typ_sim, aes(time, Cc), linewidth = 0.8) +
  annotate("rect", xmin = 0, xmax = 10, ymin = 1, ymax = Inf, alpha = 0.15) +
  facet_wrap(~arm) +
  scale_y_log10() +
  coord_cartesian(xlim = c(0, 360)) +
  labs(x = "Time (min)", y = "Morphine in blood (ng/mL)")
```

## Figure 3 / 6 -- morphine in brain ECF

```{r fig3, fig.cap = "Replicates Figures 3 and 6 of Groenendaal 2007: simulated brain ECF morphine concentrations by dose group, with the typical-value prediction and the 2.5-97.5% simulated interval. The 4 mg/kg arms reach a near-plateau; the 40 mg/kg arm declines clearly after its early peak."}
ecf_q <- ecf |>
  group_by(arm, time) |>
  summarise(lo = quantile(obs, 0.025), hi = quantile(obs, 0.975), .groups = "drop")

ggplot(ecf, aes(time, obs)) +
  geom_point(colour = "grey60", size = 0.5, alpha = 0.5) +
  geom_ribbon(data = ecf_q, aes(time, ymin = pmax(lo, 0), ymax = hi),
              inherit.aes = FALSE, alpha = 0.2) +
  geom_line(data = typ_sim, aes(time, Cbrain_ecf), linewidth = 0.8) +
  annotate("rect", xmin = 0, xmax = 10, ymin = 0, ymax = Inf, alpha = 0.15) +
  facet_wrap(~arm, scales = "free_y") +
  labs(x = "Time (min)", y = "Morphine in brain ECF (ng/mL)")
```

## Figure 4 -- dose-normalised brain ECF

Normalising to the 4 mg/kg dose makes the non-linearity explicit: if brain
distribution were linear, the three curves would superimpose.

```{r fig4, fig.cap = "Replicates Figure 4 of Groenendaal 2007: brain ECF concentrations normalised to a 4 mg/kg dose. The 40 mg/kg profile sits well below the 4 mg/kg profiles over most of the interval, the signature of a saturable active influx."}
ggplot(typ_sim, aes(time, Cbrain_ecf * 4 / dose_mgkg, colour = arm)) +
  geom_line(linewidth = 0.8) +
  annotate("rect", xmin = 0, xmax = 10, ymin = 0, ymax = Inf, alpha = 0.15) +
  labs(x = "Time (min)", y = "Dose-normalised brain ECF (ng/mL per 4 mg/kg)",
       colour = NULL) +
  theme(legend.position = "bottom")
```

## Figure 5 -- simulated dose range

```{r fig5, fig.cap = "Replicates Figure 5 of Groenendaal 2007: typical-value brain ECF profiles across the 4-40 mg/kg dose range with and without GF120918. Stable plateau concentrations at low dose give way to a clear decline as dose increases."}
dose_grid <- tidyr::crossing(dose_mgkg = c(4, 10, 20, 40), gf = c(0, 1)) |>
  mutate(id = dplyr::row_number(), WT = REF_WT, CONMED_ELACRIDAR = gf,
         amt = dose_mgkg * REF_WT * 1e6,
         label = paste0(dose_mgkg, " mg/kg"),
         pgp = ifelse(gf == 1, "+ GF120918", "vehicle"))

dg_dose <- dose_grid |>
  transmute(id, label, pgp, WT, CONMED_ELACRIDAR, time = 0, amt, rate = amt / 10,
            evid = 1L, cmt = "central", dvid = NA_integer_)
dg_obs <- dose_grid |>
  select(id, label, pgp, WT, CONMED_ELACRIDAR) |>
  tidyr::crossing(time = seq(0, 360, by = 2), dvid = 1:2) |>
  mutate(amt = NA_real_, rate = NA_real_, evid = 0L, cmt = "central")

dg_sim <- rxode2::rxSolve(
  mod, bind_rows(dg_dose, dg_obs[names(dg_dose)]) |> arrange(id, time, dvid),
  keep = c("label", "pgp"), omega = NA, sigma = NA,
  useLinCmt = FALSE, returnType = "data.frame"
) |>
  as_tibble() |>
  filter(CMT == min(CMT))

ggplot(dg_sim, aes(time, Cbrain_ecf, colour = label)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~pgp) +
  labs(x = "Time (min)", y = "Morphine in brain ECF (ng/mL)", colour = "Dose") +
  theme(legend.position = "bottom")
```

# PKNCA validation

NCA is run on the individual predictions (rather than on the residual-error
simulated observations) so the log-down trapezoid never sees a negative
concentration. The paper's comparator is a group **mean**, and the summary here
is the cohort median.

```{r pknca-setup}
dose_df <- subj |>
  transmute(id, arm, time = 0, amt)

# Blood
conc_blood <- blood |>
  filter(!is.na(ipred)) |>
  transmute(id, arm, time, Cc = ipred)

blood_obj <- PKNCA::PKNCAconc(conc_blood, Cc ~ time | arm + id,
                              concu = "ng/mL", timeu = "min")
dose_obj  <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id, doseu = "ng")

blood_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  blood_obj, dose_obj,
  intervals = data.frame(start = 0, end = Inf,
                         cmax = TRUE, tmax = TRUE, auclast = TRUE,
                         aucinf.obs = TRUE, half.life = TRUE, cl.obs = TRUE)
))
summary(blood_res)
```

```{r pknca-ecf}
conc_ecf <- ecf |>
  filter(!is.na(ipred)) |>
  transmute(id, arm, time, Cc = ipred)

ecf_obj <- PKNCA::PKNCAconc(conc_ecf, Cc ~ time | arm + id,
                            concu = "ng/mL", timeu = "min")

# The paper's brain-ECF exposure metric is AUC over the largest common
# interval, 0-165 min.
ecf_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  ecf_obj, dose_obj,
  intervals = data.frame(start = 0, end = 165,
                         cmax = TRUE, tmax = TRUE, auclast = TRUE)
))
summary(ecf_res)
```

## Blood: an exact internal identity

For a linear disposition model, AUC(0-inf) in blood must equal Dose/CL exactly,
and PKNCA's `cl.obs` must return the subject's own clearance. Both sides of this
comparison come from the *same* simulated subject -- `cl.obs` is
`Dose / AUCinf.obs` computed from that subject's profile, and `cl` is the
individual clearance the solver used to generate it -- so the only thing
separating them is trapezoidal and lambda-z error on the sampling grid. There is
no cohort randomness in the difference, which is why it is asserted tightly.

The comparison must use the **individual** clearance, not the typical value:
IIV on CL is `omega^2 = 0.129` (a ~36% CV on the log scale), so comparing NCA
clearance against the typical-value CL differs by `exp(etalcl)` and spreads over
more than a hundred percent -- a property of the cohort, not a model defect.

```{r blood-identity}
cl_cmp <- as.data.frame(blood_res) |>
  filter(PPTESTCD == "cl.obs") |>
  transmute(id = as.integer(as.character(id)), cl_nca = PPORRES) |>
  # join, rather than positional alignment, so the comparison cannot be
  # silently transposed by a change in grouping or sort order
  inner_join(distinct(blood, id, cl), by = "id") |>
  mutate(pct = 100 * (cl_nca - cl) / cl)

stopifnot(nrow(cl_cmp) == nrow(subj))

tibble(
  statistic = c("median % diff", "90th pct |% diff|", "max |% diff|"),
  value = c(median(cl_cmp$pct), quantile(abs(cl_cmp$pct), 0.9),
            max(abs(cl_cmp$pct)))
)

stopifnot(
  # Algebraic identity between PKNCA's Dose/AUCinf.obs and the solver's own
  # individual CL. Realised max |% diff| ~ 0.52% on this grid.
  max(abs(cl_cmp$pct)) < 2
)
```

## Brain ECF: comparison against the published NCA

The paper reports the dose-normalised brain ECF AUC over 0-165 min as
6810 +/- 1890, 8460 +/- 2790 and 3990 +/- 2180 for the 4 mg/kg,
4 mg/kg + GF120918 and 40 mg/kg arms respectively (Results, *PK in brain ECF*).
The printed unit is `ng h ml-1`, but the values are `ng min ml-1` -- see the
Errata below; the comparison uses the numbers as `ng*min/mL`.

```{r nca-comparison}
ecf_sim_nca <- as.data.frame(ecf_res) |>
  left_join(distinct(subj, arm, dose_mgkg), by = "arm") |>
  mutate(PPORRES = ifelse(PPTESTCD == "auclast",
                          PPORRES * 4 / dose_mgkg,   # dose-normalise to 4 mg/kg
                          PPORRES)) |>
  select(arm, PPTESTCD, PPORRES)

ecf_ref <- tibble(
  arm     = factor(levels(subj$arm), levels = levels(subj$arm)),
  auclast = c(6810, 8460, 3990)
)

nca_tbl <- nlmixr2lib::ncaComparisonTable(
  ecf_sim_nca, ecf_ref,
  by = "arm",
  params = "auclast",
  units = c(auclast = "ng*min/mL, dose-normalised to 4 mg/kg")
)
knitr::kable(nca_tbl)
attr(nca_tbl, "footnote")
```

```{r nca-assert}
auc_typ <- typ_sim |>
  filter(time <= 165) |>
  group_by(arm, dose_mgkg) |>
  summarise(
    auc = sum(diff(time) * (head(Cbrain_ecf, -1) + tail(Cbrain_ecf, -1)) / 2),
    .groups = "drop"
  ) |>
  mutate(auc_dn = auc * 4 / dose_mgkg,
         published = c(6810, 8460, 3990),
         pct = 100 * (auc_dn - published) / published)

knitr::kable(auc_typ, digits = 1)

stopifnot(
  # Typical-value prediction vs the paper's group means. Deterministic (no IIV,
  # no residual error, fixed reference weight), so this is a transcription
  # check, not a cohort statistic. Realised +0.3 / +11.5 / -8.0%. A
  # mis-transcribed kdiff, keff, N*max, C50, dose or infusion duration moves
  # these by tens of percent. The paper's own between-animal SDs on these means
  # are 28-55% of the mean, so 20% is a tight bound relative to the data.
  max(abs(auc_typ$pct)) < 20,

  # The paper's central finding: dose-normalised brain exposure FALLS with
  # dose. Deterministic typical-value comparison, not a race between two noisy
  # statistics.
  auc_typ$auc_dn[auc_typ$arm == "40 mg/kg vehicle"] <
    0.75 * auc_typ$auc_dn[auc_typ$arm == "4 mg/kg vehicle"],

  # Pgp inhibition raises brain exposure (keff falls 42%).
  auc_typ$auc_dn[auc_typ$arm == "4 mg/kg + GF120918"] >
    1.1 * auc_typ$auc_dn[auc_typ$arm == "4 mg/kg vehicle"]
)
```

## The saturable influx really is saturated

The paper's claim is that the active influx is saturated over the whole studied
range, which is what makes the dose-normalised exposure fall. With
`C50 = 9.92` ng/mL against blood concentrations of tens to tens of thousands of
ng/mL, the influx term sits near its maximum throughout.

The tightest point is the *end* of the window, not the start: blood
concentrations peak at the end of the 10 min infusion and then decline, so the
influx is least saturated at 165 min. Even there it is 85% saturated in the
4 mg/kg arms and 98% in the 40 mg/kg arm.

```{r saturation}
c50   <- exp(mod$theta[["lc50"]])
nstar <- exp(mod$theta[["lNstarMax"]])

sat <- typ_sim |>
  filter(time > 0, time <= 165) |>
  group_by(arm) |>
  summarise(
    min_blood             = min(Cc),
    t_at_min_blood        = time[which.min(Cc)],
    min_saturation_pct    = 100 * min(Cc / (c50 + Cc)),
    median_saturation_pct = 100 * median(Cc / (c50 + Cc)),
    .groups = "drop"
  )
knitr::kable(sat, digits = 2)

stopifnot(
  # All three are deterministic: typ_sim is solved with omega = NA, sigma = NA
  # at the fixed reference weight, so there is no cohort randomness here and
  # tight bounds are appropriate.

  # The influx term never drops below 80% of N*max anywhere in the 0-165 min
  # window, so it acts as a near-constant input. Realised minimum 84.9%
  # (4 mg/kg arms, at t = 165 min).
  all(sat$min_saturation_pct > 80),

  # Typical saturation across the window is far higher still. Realised 93.4%
  # (4 mg/kg) and 99.3% (40 mg/kg).
  all(sat$median_saturation_pct > 90),

  # The mechanism: the 10-fold higher dose sits even deeper into saturation,
  # so the influx contributes proportionally LESS at 40 mg/kg -- which is what
  # drives the dose-normalised brain exposure down.
  sat$min_saturation_pct[sat$arm == "40 mg/kg vehicle"] >
    sat$min_saturation_pct[sat$arm == "4 mg/kg vehicle"]
)
```

## Efflux inhibition magnitude

```{r efflux}
keff_veh <- exp(mod$theta[["lkefflux_bbb"]])
keff_gf  <- exp(mod$theta[["lkefflux_bbb"]] +
                mod$theta[["e_conmed_elacridar_kefflux_bbb"]])

tibble(
  parameter = c("keff, vehicle (1/min)", "keff, + GF120918 (1/min)",
                "reduction (%)"),
  value = c(keff_veh, keff_gf, 100 * (1 - keff_gf / keff_veh))
)

stopifnot(
  abs(keff_veh - 0.0195) < 1e-9,
  abs(keff_gf  - 0.0113) < 1e-9,
  # Abstract: "active efflux, reduced by 42% by Pgp inhibition".
  abs(100 * (1 - keff_gf / keff_veh) - 42) < 1
)
```

# Assumptions and deviations

* **`median BW` is not printed.** Equation 7 centres the body-weight effect on
  the cohort median, but the paper never gives the number. 0.300 kg is used.
  Three independent lines support it: the stated weight range is 250-350 g
  (midpoint 300 g); the per-group means in Table 1 span 0.260-0.306 kg and the
  cohort-weighted mean is 0.294 kg; the largest single group (N = 20) has a mean
  of exactly 0.300 kg. Moving the reference to 0.294 kg shifts typical CL by
  about 3% and typical V2 by about 5%.
* **Body weights are drawn uniformly over 250-350 g.** The paper reports only
  the range and per-group means, not individual weights or a distribution.
* **The two model layers were fitted sequentially, but are simulated jointly
  here.** The paper fitted the blood model first and then used it as the input
  function for the brain ODE. The packaged model integrates the blood and brain
  systems together, which is the same system of equations; there is no
  back-coupling term from brain to blood in either formulation (brain uptake is
  negligible relative to systemic disposition).
* **`brain_ecf` is a concentration state.** Paper equations 10-11 aggregate
  every brain parameter by dividing by the unidentified $V_{ecf}$, so an
  amount-based state is not reconstructable from the published values.
* **The residual-error rows are read as variances.** Paper equations 5-6 define
  `eps ~ N(0, sigma^2)`, and the Table 2 / Table 3 "Proportional error" rows are
  those `sigma^2` values, so `propSd = sqrt(0.074) = 0.2720` and
  `propSd_Cbrain_ecf = sqrt(0.094) = 0.3066`. The sibling
  `Geldof_2008_fluvoxamine_rat` (same Leiden group, shared author J Freijer,
  identical notation) reports `sigma^2 = 0.042` explicitly and is encoded the
  same way. Reading these rows as SDs would imply a 7.4% residual on in-vivo rat
  blood data, below the assay's own inter-assay variability.
* **The GF120918 arm switch is re-encoded multiplicatively.** Paper equation 8
  estimates two independent `keff` values; the packaged model uses the canonical
  `exp(lkefflux_bbb + e_conmed_elacridar_kefflux_bbb * CONMED_ELACRIDAR)` form,
  which is numerically identical at both covariate levels.
* **IIV on all parameters other than CL, V2, kdiff and keff was fixed to zero
  by the authors**, not by this extraction (paper Results).
* **Brain-ECF Tmax in the GF120918 arm.** The paper states the ECF maximum was
  reached "within 40 min" for the 4 mg/kg groups. The model's typical-value
  GF120918 profile peaks nearer 85 min. The profile is very flat there (the
  predicted 165 min concentration is within 10% of the peak), and the
  observations are 15 min dialysate fractions carrying about 31% residual error,
  so an observed Tmax anywhere in that window is consistent. No assertion is
  placed on Tmax for that arm.
* **The 10 mg/kg arm is simulated only implicitly.** It contributed blood data
  to Table 2 but had no microdialysis animals, so it is not one of the three
  brain-ECF arms.

# Errata

Discrepancies found in the published paper. None changes a point estimate used
by the model except where noted.

1. **Table 2, Slope factor confidence intervals are transposed.** The Cl slope
   factor is 5.35 with a printed CI of 5.66-11.3, and the V2 slope factor is
   8.50 with a printed CI of 2.70-7.99 -- each point estimate falls *outside*
   its own printed interval and *inside* the other's. Reconstructing the
   intervals as `estimate +/- 1.96 * CV% * estimate` gives 2.71-7.99 for the Cl
   slope and 5.65-11.35 for the V2 slope, i.e. the two cells are swapped. Point
   estimates are unaffected.
2. **Table 3, N\*max units.** Printed as `ng min-1`. Equation 10 places
   `Nmax/Vecf` additively against `kdiff * (Cblood - Cecf)` inside `dCecf/dt`,
   so it must be `ng mL-1 min-1`. The Abstract states it correctly as
   "(Nmax/Vecf) of 0.66 ng min-1 ml-1". The model uses ng/mL/min.
3. **Results, dose-normalised AUC units.** The values 6810 / 8460 / 3990 are
   printed as `ng h ml-1` but are `ng min ml-1`. The observed brain ECF
   concentrations in Figure 3 are tens of ng/mL for the 4 mg/kg arms, so an AUC
   of 6810 ng*h/mL over 2.75 h would require a mean concentration of about
   2480 ng/mL -- roughly 50-fold above anything measured. Read as ng*min/mL the
   published values are reproduced by the model to within 12% (table above).
4. **Three confidence-interval lower limits lost their minus sign** to the PDF's
   symbol font: Table 3 `omega^2 kdiff` prints 0.025-0.501 for
   `0.238 +/- 1.96 * 56.3% * 0.238 = (-0.025, 0.501)`; the kdiff-keff covariance
   prints 0.039-0.158 for `(-0.039, 0.158)`; and `C50` prints 3.97-23.8 for
   `(-3.98, 23.8)`. Every other interval in Tables 2 and 3 reproduces exactly
   under the same formula, which is what identifies these three.
5. **Table 3, the covariance row label is mangled** as `o2keffB o2keff`. The
   Results text states a covariance block was added because the post-hoc
   individual `kdiff` and `keff` estimates correlated, so the row is
   `cov(eta_kdiff, eta_keff) = 0.059`. Read as a covariance it gives a
   correlation of `0.059 / sqrt(0.238 * 0.080) = 0.428`; the CI reconstruction in
   item 4 confirms 0.059 is the tabulated estimate.
6. **Table 2, Cl Intercept confidence interval.** Printed 18.5-23.1;
   `20.0 +/- 1.96 * 5.6% * 20.0` gives 17.8-22.2. Every other Table 2 interval
   reconstructs exactly, so this one cell appears mis-set. The point estimate is
   unaffected.
7. **Efflux reduction, 40% vs 42%.** The Results text says `keff` "decreased by
   about 40%"; the Abstract says 42%. `1 - 0.0113/0.0195 = 42.05%`, so the
   Abstract is right and the Results text is rounded.
8. **GF120918 maintenance infusion rate.** Printed as a 25 ng/min continuous
   infusion following the 6 mg/kg bolus. That rate is hard to reconcile with the
   measured mean steady-state blood concentration of 214 ng/mL. It is
   transcribed as printed in the model metadata; it is narrative only and does
   not enter the model, which carries GF120918 as a binary indicator.

9. **The MD-only group's treatment arm is self-contradictory.** *Experimental
   procedures* states "The MD rats all received a 10-min infusion of
   4 mg kg-1 morphine and a continuous infusion of GF120918", but Table 1 lists
   the MD group (N = 3) under Treatment = "Vehicle". The two statements cannot
   both hold. This affects only how 3 of the 32 brain-ECF animals are allocated
   between the 4 mg/kg vehicle and 4 mg/kg + GF120918 arms; it changes no
   parameter estimate, and the packaged model carries the arm as a covariate
   rather than a fixed cohort composition. The population metadata follows
   Table 1, because Table 1 is the table the paper's own group-level N counts
   and body weights are keyed to.

No erratum or corrigendum is attached to this article: a Crossref query on
`10.1038/sj.bjp.0707257` returns no `update-to` or `updated-by` relation.

# Session info

```{r session}
sessionInfo()
```
