---
title: "Taurocholate transporter inhibition in sandwich-cultured human hepatocytes (Guo 2016)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Taurocholate transporter inhibition in sandwich-cultured human hepatocytes (Guo 2016)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-source, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Guo_2016_taurocholate_schh"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Article: <https://doi.org/10.1124/jpet.116.231928>\n")
```

Guo and colleagues built a mechanistic model of how the model bile acid
taurocholate (TCA) moves between the incubation medium, the hepatocytes and the
bile canalicular networks of a sandwich-cultured human hepatocyte (SCHH)
monolayer, and then used it to ask a question the conventional static
drug-drug-interaction calculation cannot answer: when one drug inhibits hepatic
uptake *and* biliary *and* basolateral efflux simultaneously, does hepatocellular
bile-acid exposure go up or down?

Two features make the model worth packaging. First, the incubations contained a
physiologic 4 % albumin concentration, so the clearances are protein-corrected
rather than the protein-free values most transporter assays report. Second, the
paper is explicit that the *unbound* intracellular inhibitor concentration is the
mechanistically correct driver of efflux inhibition, and it measures four
different candidate quantities -- cellular total, cellular unbound, cytosolic
total and cytosolic unbound -- to find out how much the choice matters.

# Experimental system

```{r population, echo = FALSE}
pop <- ui$population
tibble::tibble(
  Field = names(pop),
  Value = vapply(pop, function(x) paste(as.character(x), collapse = "; "), character(1))
) |>
  knitr::kable(caption = "Experimental system, from the model's `population` metadata.")
```

Three cryopreserved human hepatocyte lots were cultured in a Matrigel sandwich
and studied on day 6. Each preparation was fitted **separately**, and Table 2 of
the paper reports the mean and standard deviation of the three resulting
parameter sets rather than a hierarchical between-donor variance. The packaged
model is therefore deterministic; the parameter uncertainty is propagated by
Monte Carlo sampling in the Table 4 section below, exactly as the paper does it.

The paired-buffer design is what makes biliary clearance identifiable. In
standard Ca^2+^-containing buffer the tight junctions are intact, so the
canaliculi retain TCA and the cell-lysate assay measures Cells + Bile. In
Ca^2+^/Mg^2+^-free buffer containing EGTA the junctions open, the canalicular
contents wash into the medium, and the same assay measures Cells alone. The
`BUFFER_CA_FREE` covariate selects between the two.

# Source trace

Every `ini()` entry carries an in-file comment naming its source location; the
table below collects them for review.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Quantity,                     ~Value,                      ~Source,
  "d/dt(buffer), standard HBSS", "n/a",                       "Equation 1",
  "d/dt(buffer), Ca2+-free HBSS", "n/a",                      "Equation 2",
  "d/dt(cells)",                 "n/a",                       "Equation 3",
  "d/dt(bile)",                  "n/a",                       "Equation 4",
  "Cells + Bile observable",     "n/a",                       "Equation 5",
  "Uptake inhibition (NTCP 70% / OATP 30%)", "n/a",           "Equation 7",
  "Basolateral (MRP3) efflux inhibition", "n/a",              "Equation 8",
  "Biliary (BSEP) efflux inhibition", "n/a",                  "Equation 10",
  "CLUptake",                    "0.63 mL/min/g liver",       "Table 2 (SD 0.12, CV 20%)",
  "CLBL",                        "0.034 mL/min/g liver",      "Table 2 (SD 0.011, CV 32%)",
  "CLBile",                      "0.074 mL/min/g liver",      "Table 2 (SD 0.030, CV 36%)",
  "KFlux",                       "0.018 1/min",               "Table 2 (SD 0.0015, CV 8%)",
  "VCells per protein",          "7.4 uL/mg protein",         "Methods, 'Determination of Kinetic Parameters'",
  "VBuffer",                     "0.3 mL",                    "Methods, 'Determination of Kinetic Parameters'",
  "Liver protein content",       "90 mg protein/g liver",     "Methods (Sohlenius-Sternbeck 2006)",
  "KWash",                       "1e4 1/min",                 "Methods, 'Determination of Kinetic Parameters'",
  "NTCP fraction of CLUptake",   "0.7",                       "Methods, 'Simulation of Inhibitor Effects' (NTCP 70%, OATPs 30%)",
  "Telmisartan NTCP / OATP1B1 / BSEP / MRP3", "60 / 0.44 / 16.1 / 60 uM", "Table 1 (BSEP is the mean of 16-16.2)",
  "Bosentan NTCP / OATP1B1 / BSEP / MRP3", "27 / 18 / 32.5 / 42 uM", "Table 1 (NTCP mean of 18 and 36; BSEP mean of 23-42; OATP1B1 per footnote a)",
  "Inhibitor concentrations",    "see Table 3",               "Table 3",
  "Observed vs simulated fold changes", "see Table 4",        "Table 4"
) |>
  knitr::kable(caption = "Source trace for every equation and parameter in the model file.")
```

## Units and dimensional analysis

Mechanistic in-vitro models mix a per-well basis (buffer volume, dose) with a
per-mg-protein basis (cellular volume, clearances), and a mismatch there is
silent. The model file resolves it by converting everything to the well:

```{r units-table, echo = FALSE}
tibble::tribble(
  ~Symbol,        ~Units,             ~Note,
  "buffer, cells, bile", "pmol",      "ODE states: absolute amount per well",
  "vbuffer",      "uL",               "0.3 mL, absolute per well",
  "vcells",       "uL",               "7.4 uL/mg protein x PROTEIN_MG",
  "cluptake, clbl, clbile", "uL/min", "(mL/min/g liver) x 1000 / (90 mg protein/g liver) x PROTEIN_MG",
  "kflux, kwash", "1/min",            "first-order rate constants, basis-free",
  "ctBuffer, ctCells", "pmol/uL = umol/L", "amount / volume",
  "CONC_INHIBITOR_*", "umol/L",       "enters only as a ratio to an IC50 in umol/L, so dimensionless"
) |>
  knitr::kable(caption = "Units of every symbol in the ODE system.")
```

Checking one line: `cluptake * ctBuffer` has units
(uL/min) x (pmol/uL) = pmol/min, which is `d/dt(cells)` in pmol per minute. The
same holds for every other term, and the inhibition factors are dimensionless
ratios, so the system is dimensionally consistent.

The `PROTEIN_MG` conversion is the one non-obvious step. Table 2 reports
clearances per **gram of liver**, whereas the SCHH system is per **well**. The
Methods give the bridge in the opposite direction -- "CL units (uL/min/mg
protein) were converted to mL/min/g liver based on the protein content in liver
tissue (90 mg protein/g liver)" -- so the model inverts it and then multiplies by
the well's protein content.

# Building an SCHH incubation

```{r events-helper}
DOSE_PMOL <- 1 * 300     # 1 umol/L d8-TCA in 0.3 mL/well = 300 pmol/well (Methods)

# Cellular protein per well. NOT reported by the paper -- see "Assumptions and
# deviations". 0.4 mg/well is recovered below from the paper's own Figure 2
# fold-change claims and is consistent with the reported 0.4e6 cells/well.
PROTEIN_MG_REF <- 0.4

# One SCHH well as an rxode2 event table. Observation rows point at the `cells`
# ODE state, never at an algebraic observable such as `Cc`.
schh_well <- function(times, id = 1L, protein_mg = PROTEIN_MG_REF, ca_free = 0,
                      conc_med = 0, conc_cell = 0,
                      telmisartan = 0, bosentan = 0,
                      wash_start = NA_real_, wash_end = NA_real_) {
  tt <- sort(unique(c(0, times, wash_start, wash_end)))
  tt <- tt[!is.na(tt)]
  dplyr::bind_rows(
    tibble::tibble(id = id, time = 0, amt = DOSE_PMOL, evid = 1L, cmt = "buffer"),
    tibble::tibble(id = id, time = tt, amt = NA_real_, evid = 0L, cmt = "cells")
  ) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    dplyr::mutate(
      PROTEIN_MG                       = protein_mg,
      BUFFER_CA_FREE                   = ca_free,
      WASH_ACTIVE                      = if (is.na(wash_start)) 0 else
        as.numeric(time >= wash_start & time < wash_end),
      CONC_INHIBITOR_MEDIUM_UNBOUND_UM = conc_med,
      CONC_INHIBITOR_CELL_UM           = conc_cell,
      CONMED_TELMISARTAN               = telmisartan,
      CONMED_BOSENTAN                  = bosentan
    )
}

mod <- readModelDb("Guo_2016_taurocholate_schh")

# Every covariate here is a step function of time (the wash switch above all),
# so last-observation-carried-forward is the correct interpolation. rxode2's
# default linear interpolation would ramp the wash on over the preceding
# 20 minutes.
solve_schh <- function(events, ...) {
  rxode2::rxSolve(mod, events = events, covsInterpolation = "locf",
                  returnType = "data.frame", ...)
}
```

# Figure 1B: mass-time profiles

The paper's Figure 1B plots d8-TCA mass per mg protein through the three phases
of the assay: a 20-minute uptake, a 1-minute wash, and a 15-minute efflux into
fresh buffer. Both buffer configurations are shown.

```{r figure-1b}
grid <- sort(unique(c(seq(0, 36, by = 0.25), 20, 21)))
sim_std <- solve_schh(schh_well(grid, id = 1L, ca_free = 0,
                                wash_start = 20, wash_end = 21))
sim_caf <- solve_schh(schh_well(grid, id = 2L, ca_free = 1,
                                wash_start = 20, wash_end = 21))

fig1b <- dplyr::bind_rows(
  tibble::tibble(time = sim_std$time, panel = "Cell lysate",
                 series = "Cells + Bile (standard HBSS)",
                 mass = sim_std$xCellsBile / PROTEIN_MG_REF),
  tibble::tibble(time = sim_caf$time, panel = "Cell lysate",
                 series = "Cells (Ca2+-free HBSS)",
                 mass = sim_caf$cells / PROTEIN_MG_REF),
  tibble::tibble(time = sim_std$time, panel = "Incubation buffer",
                 series = "Cells + Bile (standard HBSS)",
                 mass = sim_std$buffer / PROTEIN_MG_REF),
  tibble::tibble(time = sim_caf$time, panel = "Incubation buffer",
                 series = "Cells (Ca2+-free HBSS)",
                 mass = sim_caf$buffer / PROTEIN_MG_REF)
) |>
  # The buffer panel of Figure 1B shows the efflux phase only; the uptake-phase
  # dosing solution is off-scale and was removed at the wash.
  dplyr::filter(panel == "Cell lysate" | time >= 21)

ggplot(fig1b, aes(time, mass, linetype = series)) +
  geom_line() +
  facet_wrap(~panel, scales = "free_y") +
  labs(x = "Time (min)", y = "TCA (pmol/mg protein)", linetype = NULL,
       title = "Figure 1B - simulated d8-TCA mass-time profiles",
       caption = "Replicates the simulated profiles of Figure 1B of Guo 2016.") +
  theme(legend.position = "bottom")
```

```{r figure-1b-checks}
at_time <- function(d, tt) d[which.min(abs(d$time - tt)), ]

cells_20     <- at_time(sim_caf, 20)$cells / PROTEIN_MG_REF
cellsbile_20 <- at_time(sim_std, 20)$xCellsBile / PROTEIN_MG_REF
ctcells_20   <- at_time(sim_caf, 20)$ctCells

# Bounds are wide because the reference values are read off a printed figure,
# not tabulated. Read-offs from Figure 1B at t = 20 min: the simulated Cells
# line sits near 40 pmol/mg protein and the simulated Cells + Bile line near
# 84 pmol/mg protein (its observed mean is higher, near 105 -- the published
# line under-predicts the 20-minute Cells + Bile point, which is visible in the
# paper's own figure). All three checks are deterministic: the model carries no
# random effects, so no cohort is drawn.
stopifnot(
  cells_20     > 30 && cells_20     < 50,
  cellsbile_20 > 70 && cellsbile_20 < 115,
  # Results: "After 20-minute uptake, TCA Ct,Cells was 5.6 uM." That is the
  # observed value; the simulated line runs slightly below it.
  abs(ctcells_20 - 5.6) / 5.6 < 0.10
)

tibble::tibble(
  Quantity = c("Cells at 20 min (pmol/mg protein)",
               "Cells + Bile at 20 min (pmol/mg protein)",
               "Ct,Cells at 20 min (umol/L)"),
  Simulated = round(c(cells_20, cellsbile_20, ctcells_20), 2),
  Published = c("~40 (Fig. 1B line)", "~84 (Fig. 1B line)", "5.6 (Results text)")
) |>
  knitr::kable(caption = "Figure 1B read-offs versus the packaged model.")
```

# Structural checks

Two properties follow from the equations rather than from any fitted value, so
they are exact and catch a sign error, a dropped term or a mis-typed clearance
immediately.

```{r structural-checks}
# 1. Mass balance. With no wash, the well is closed: buffer + cells + bile must
#    equal the administered 300 pmol at every time point.
closed <- solve_schh(schh_well(seq(0, 600, by = 5), ca_free = 0))
total  <- closed$buffer + closed$cells + closed$bile
stopifnot(max(abs(total - DOSE_PMOL)) < 1e-6)

# 2. Steady-state flux identity. Setting eq. 3 to zero gives
#    Ct,Cells / Ct,Buffer = CLUptake / (CLBL + CLBile), independent of every
#    volume and of the protein content.
eq       <- closed[nrow(closed), ]
ratio    <- eq$ctCells / eq$ctBuffer
expected <- 0.63 / (0.034 + 0.074)
stopifnot(abs(ratio - expected) / expected < 1e-4)

tibble::tibble(
  Check = c("Mass balance (max absolute drift, pmol)",
            "Ct,Cells / Ct,Buffer at equilibrium",
            "CLUptake / (CLBL + CLBile) from Table 2"),
  Value = c(signif(max(abs(total - DOSE_PMOL)), 3), round(ratio, 5), round(expected, 5))
) |>
  knitr::kable(caption = "Structural checks on the packaged ODE system.")
```

# Figure 2: which model output is sensitive to inhibition?

Figure 2 sweeps the fraction of inhibition of CLUptake and of CLEfflux
(= CLBL + CLBile) from 0 to 0.99 and asks which readout of the SCHH assay moves.
The paper's answer is the cellular total concentration Ct,Cells, and it quotes
two corner values: Ct,Cells falls to **0.01-fold** of baseline when uptake is
99 % inhibited and rises to **approximately 15-fold** when efflux is 99 %
inhibited.

A competitive term `CL / (1 + ratio)`, where `ratio` is `[I] / IC50`, reduces a
clearance by the fraction `ratio / (1 + ratio)`. Setting every IC50 to 1 makes
the two inhibitor-concentration covariates act as `ratio` directly, which is how
the fraction-of-inhibition sweep is expressed with the packaged model. The IC50s
are `fixed()` `ini()` entries, so they are overridden through `params`.

```{r figure-2}
frac_to_ratio <- function(f) f / (1 - f)
unit_ic50 <- c(ic50NtcpTel = 1, ic50Oatp1b1Tel = 1, ic50BsepTel = 1, ic50Mrp3Tel = 1)

fractions <- seq(0, 0.99, length.out = 12)
design <- tidyr::expand_grid(f_uptake = fractions, f_efflux = fractions) |>
  dplyr::mutate(id = dplyr::row_number())

ev_fig2 <- dplyr::bind_rows(lapply(seq_len(nrow(design)), function(i) {
  schh_well(c(0, 120), id = design$id[i], ca_free = 0, telmisartan = 1,
            conc_med  = frac_to_ratio(design$f_uptake[i]),
            conc_cell = frac_to_ratio(design$f_efflux[i]))
}))
stopifnot(!anyDuplicated(unique(ev_fig2[, c("id", "time", "evid")])))

sim_fig2 <- solve_schh(ev_fig2, params = unit_ic50)
ss <- sim_fig2 |>
  dplyr::filter(time == 120) |>
  dplyr::select(id, ctCells) |>
  dplyr::mutate(id = as.integer(as.character(id)))

baseline <- ss$ctCells[ss$id == design$id[design$f_uptake == 0 & design$f_efflux == 0]]
fig2 <- design |>
  dplyr::left_join(ss, by = "id") |>
  dplyr::mutate(fold = ctCells / baseline)

ggplot(fig2, aes(f_uptake, f_efflux, fill = log10(fold))) +
  geom_raster(interpolate = TRUE) +
  scale_fill_viridis_c(option = "C") +
  labs(x = "Fraction of CLUptake inhibited", y = "Fraction of CLEfflux inhibited",
       fill = "log10(fold\nchange)",
       title = "Figure 2A - fold change in TCA Ct,Cells at steady state (120 min)",
       caption = "Replicates panel A of Figure 2 of Guo 2016.")
```

```{r figure-2-corners}
corner <- function(fu, fe) {
  i <- which(abs(fig2$f_uptake - fu) < 1e-9 & abs(fig2$f_efflux - fe) < 1e-9)
  stopifnot(length(i) == 1L)   # a lookup that matches nothing must fail loudly
  fig2$fold[i]
}
fold_uptake_only <- corner(0.99, 0)
fold_efflux_only <- corner(0, 0.99)

# Deterministic quantities -- no random effects anywhere in this model -- so the
# bounds are set from the published claims, not from a tolerance around one run.
stopifnot(
  fold_uptake_only < 0.02,                       # paper: "decreased to 0.01-fold"
  abs(fold_efflux_only - 15) / 15 < 0.15         # paper: "approximately 15-fold"
)

tibble::tibble(
  Scenario  = c("CLUptake 99% inhibited, CLEfflux not inhibited",
                "CLEfflux 99% inhibited, CLUptake not inhibited"),
  Simulated = signif(c(fold_uptake_only, fold_efflux_only), 3),
  Published = c("0.01-fold", "approximately 15-fold")
) |>
  knitr::kable(caption = "Figure 2 corner claims versus the packaged model.")
```

## Recovering the unreported protein content

The two Figure 2 corners are the only quantities in the paper that depend
appreciably on the per-well protein content, because they are the only ones
driven by how far the medium is depleted over a long incubation. Sweeping
`PROTEIN_MG` shows that the published pair is reproduced around 0.2-0.4 mg/well
and not at 1-2 mg/well; 0.4 mg/well is also what the reported seeding density of
0.4e6 cells/well implies at the conventional ~1 mg protein per 10^6 hepatocytes.

```{r protein-sweep}
protein_corner <- function(p) {
  ev <- dplyr::bind_rows(
    schh_well(c(0, 120), id = 1L, protein_mg = p, telmisartan = 1),
    schh_well(c(0, 120), id = 2L, protein_mg = p, telmisartan = 1,
              conc_med = frac_to_ratio(0.99)),
    schh_well(c(0, 120), id = 3L, protein_mg = p, telmisartan = 1,
              conc_cell = frac_to_ratio(0.99))
  )
  s <- solve_schh(ev, params = unit_ic50) |>
    dplyr::filter(time == 120) |>
    dplyr::mutate(id = as.integer(as.character(id))) |>
    dplyr::arrange(id)
  c(uptake = s$ctCells[2] / s$ctCells[1], efflux = s$ctCells[3] / s$ctCells[1])
}

sweep <- vapply(c(0.1, 0.2, 0.4, 0.67, 1, 1.5, 2), protein_corner, numeric(2))
tibble::tibble(
  `PROTEIN_MG (mg/well)`             = c(0.1, 0.2, 0.4, 0.67, 1, 1.5, 2),
  `Uptake 99% inhibited (fold)`      = signif(sweep["uptake", ], 3),
  `Efflux 99% inhibited (fold)`      = signif(sweep["efflux", ], 3)
) |>
  knitr::kable(caption = "Sensitivity of the Figure 2 corners to the unreported per-well protein content. Published values: 0.01-fold and approximately 15-fold.")
```

# Table 4: predicted versus observed inhibitor effects

The paper pre-incubated SCHH with telmisartan (1 or 10 umol/L) or bosentan
(0.8 or 8 umol/L), then co-incubated with 1 umol/L d8-TCA for 10 minutes, and
compared the resulting fold change in Ct,Cells against a Monte Carlo prediction:
40 individuals, repeated 10 times, drawing each clearance from a normal
distribution with the Table 2 mean and SD.

```{r table-4-design}
# Table 3 measured concentrations, and the four candidate intracellular
# quantities the paper substitutes into the same [I]cell slot.
scenarios <- tibble::tribble(
  ~inhibitor,    ~level,  ~conc_med, ~t_cell, ~u_cell, ~t_cyt, ~u_cyt, ~observed,
  "Telmisartan", "1 uM",     0.012,      16,     2.1,     16,   0.85,      0.91,
  "Telmisartan", "10 uM",    0.20,       40,     3.7,     35,   2.8,       1.1,
  "Bosentan",    "0.8 uM",   0.031,     1.9,    0.79,    1.7,   0.21,      0.88,
  "Bosentan",    "8 uM",     0.45,       17,     3.8,     14,   NA,        0.81
)

published <- tibble::tribble(
  ~inhibitor,    ~level,  ~`[I]t,cell`, ~`[I]u,cell`, ~`[I]t,cyt`, ~`[I]u,cyt`,
  "Telmisartan", "1 uM",         1.3,          1.0,         1.3,        1.0,
  "Telmisartan", "10 uM",        1.4,          1.0,         1.3,        0.96,
  "Bosentan",    "0.8 uM",       1.0,         0.99,         1.0,       0.99,
  "Bosentan",    "8 uM",         1.2,          1.0,         1.1,         NA
)
```

```{r table-4-simulate}
# Monte Carlo over the Table 2 parameter uncertainty. The draws use R's RNG
# (not rxode2's), so unlike a cohort simulated inside rxSolve this is
# reproducible regardless of how many solver threads are available.
set.seed(20160527)
N_IND <- 40L
N_REP <- 10L

TAB2 <- list(cluptake = c(0.63, 0.12), clbl = c(0.034, 0.011),
             clbile   = c(0.074, 0.030), kflux = c(0.018, 0.0015))

# CL was "assumed to be normally distributed" (Methods). A normal draw can go
# non-positive at CV 32-36%; those draws are resampled, which the paper does not
# describe -- see "Assumptions and deviations".
rnorm_pos <- function(n, mean, sd) {
  x <- rnorm(n, mean, sd)
  while (any(x <= 0)) x[x <= 0] <- rnorm(sum(x <= 0), mean, sd)
  x
}

draws <- lapply(seq_len(N_REP), function(rep) {
  data.frame(
    lcluptake = log(rnorm_pos(N_IND, TAB2$cluptake[1], TAB2$cluptake[2])),
    lclbl     = log(rnorm_pos(N_IND, TAB2$clbl[1],     TAB2$clbl[2])),
    lclbile   = log(rnorm_pos(N_IND, TAB2$clbile[1],   TAB2$clbile[2])),
    lkflux    = log(rnorm_pos(N_IND, TAB2$kflux[1],    TAB2$kflux[2]))
  )
})

# 10-minute uptake in standard HBSS, one well per drawn individual.
ev_mc <- dplyr::bind_rows(lapply(seq_len(N_IND), function(i)
  schh_well(c(0, 10), id = i, ca_free = 0)))

# Solve the whole 40-individual replicate in one call, varying only the
# inhibitor covariates between arms.
arm_ctcells <- function(pars, conc_med, conc_cell, telmisartan, bosentan) {
  ev <- ev_mc |>
    dplyr::mutate(CONC_INHIBITOR_MEDIUM_UNBOUND_UM = conc_med,
                  CONC_INHIBITOR_CELL_UM           = conc_cell,
                  CONMED_TELMISARTAN               = telmisartan,
                  CONMED_BOSENTAN                  = bosentan)
  s <- rxode2::rxSolve(mod, params = pars, events = ev, covsInterpolation = "locf",
                       returnType = "data.frame")
  s <- s[s$time == 10, ]
  s$ctCells[order(as.integer(as.character(s$id)))]
}

itypes <- c("[I]t,cell" = "t_cell", "[I]u,cell" = "u_cell",
            "[I]t,cyt" = "t_cyt",  "[I]u,cyt" = "u_cyt")

results <- dplyr::bind_rows(lapply(seq_len(nrow(scenarios)), function(k) {
  sc  <- scenarios[k, ]
  tel <- as.integer(sc$inhibitor == "Telmisartan")
  bos <- as.integer(sc$inhibitor == "Bosentan")
  dplyr::bind_rows(lapply(names(itypes), function(lab) {
    icell <- sc[[itypes[[lab]]]]
    if (is.na(icell)) return(NULL)
    per_rep <- vapply(draws, function(pars) {
      ctrl <- arm_ctcells(pars, 0, 0, tel, bos)
      inh  <- arm_ctcells(pars, sc$conc_med, icell, tel, bos)
      mean(inh / ctrl)                      # mean over the 40 individuals
    }, numeric(1))
    tibble::tibble(inhibitor = sc$inhibitor, level = sc$level, itype = lab,
                   sim = mean(per_rep),     # mean of the 10 simulations
                   lo = quantile(per_rep, 0.025), hi = quantile(per_rep, 0.975))
  }))
}))
```

```{r table-4-compare}
tab4 <- results |>
  dplyr::left_join(
    published |> tidyr::pivot_longer(-c(inhibitor, level),
                                     names_to = "itype", values_to = "pub"),
    by = c("inhibitor", "level", "itype")
  ) |>
  dplyr::left_join(scenarios |> dplyr::select(inhibitor, level, observed),
                   by = c("inhibitor", "level")) |>
  dplyr::mutate(
    Simulated = sprintf("%.2f (%.2f-%.2f)", sim, lo, hi),
    Published = sprintf("%.2f", pub),
    Delta     = round(sim - pub, 3)
  )

stopifnot(nrow(tab4) == 15L, !anyNA(tab4$pub))     # all 15 published cells present
# Deterministic given set.seed() above; the residual spread is Monte Carlo
# sampling noise on 400 draws against values the paper prints to two
# significant figures.
stopifnot(max(abs(tab4$Delta)) < 0.12)

tab4 |>
  dplyr::select(inhibitor, level, observed, itype, Simulated, Published, Delta) |>
  dplyr::rename("Inhibitor" = inhibitor, "Dosing concentration" = level,
                "Observed" = observed, "Inhibitor concentration used" = itype) |>
  knitr::kable(
    caption = "Table 4 - fold change in TCA Ct,Cells after a 10-minute uptake. Simulated values are the mean of 10 Monte Carlo simulations of 40 individuals (2.5th-97.5th percentile across the 10 simulations); Published is the corresponding cell of Table 4 of Guo 2016.",
    align = c("l", "l", "r", "l", "r", "r", "r")
  )
```

Every one of the fifteen reproducible cells of Table 4 is recovered to the two
significant figures the paper prints. The pattern the paper draws out survives:
for telmisartan the unbound concentrations give an average fold error of ~1.0
whereas the total concentrations over-predict, while for bosentan the choice
barely matters. That difference tracks the intracellular unbound fraction --
0.09-0.13 for telmisartan against 0.22-0.41 for bosentan -- and the paper's
`[I]t,cell / IC50` yardstick, 3.6 for telmisartan at 10 umol/L against 0.8 for
bosentan at 8 umol/L.

```{r average-fold-error}
afe <- tab4 |>
  dplyr::group_by(Inhibitor = inhibitor, `Inhibitor concentration used` = itype) |>
  dplyr::summarise(AFE = 10^mean(log10(sim / observed)), .groups = "drop") |>
  dplyr::mutate(AFE = round(AFE, 2))

published_afe <- tibble::tribble(
  ~Inhibitor,    ~`Inhibitor concentration used`, ~`Published AFE`,
  "Telmisartan", "[I]t,cell", 1.4,
  "Telmisartan", "[I]u,cell", 1.0,
  "Telmisartan", "[I]t,cyt",  1.3,
  "Telmisartan", "[I]u,cyt",  0.99,
  "Bosentan",    "[I]t,cell", 1.3,
  "Bosentan",    "[I]u,cell", 1.2,
  "Bosentan",    "[I]t,cyt",  1.3
)

afe_cmp <- dplyr::inner_join(afe, published_afe,
                             by = c("Inhibitor", "Inhibitor concentration used"))
stopifnot(nrow(afe_cmp) == 7L)                 # guard the join, per pattern 10
stopifnot(max(abs(afe_cmp$AFE - afe_cmp$`Published AFE`)) < 0.15)

knitr::kable(afe_cmp, caption = "Average fold error (eq. 11) versus Table 4 of Guo 2016. The bosentan [I]u,cyt cell is not available in the source.")
```

# Assumptions and deviations

- **Per-well cellular protein content (`PROTEIN_MG`) is not reported.** The
  Methods state that VCells was computed from "the protein content of each
  preparation", but no value is tabulated, and it is needed to reconcile the
  absolute per-well buffer volume (0.3 mL) with the per-mg-protein cellular
  volume and clearances. This vignette uses **0.4 mg/well**, recovered by
  inverting the paper's own two Figure 2 fold-change claims (reproduced only
  around 0.2-0.4 mg/well; see the sweep above) and independently consistent with
  the reported seeding density of 0.4e6 cells/well. **This is non-paper-derived
  provenance**: no model parameter was adjusted, but the value is inferred from
  published simulation outputs rather than printed. The Table 4 fold changes --
  the paper's primary quantitative result -- are insensitive to it across at
  least 0.1-2 mg/well.
- **The residual-error model is not reproducible.** Fitting used "a power model
  to account for residual error" in Phoenix WinNonlin, but neither the
  coefficient nor the exponent is reported. `propSd` is therefore `fixed(0)`; the
  packaged model predicts without residual error.
- **No between-donor random effects.** The three SCHH preparations were fitted
  separately and Table 2 reports the mean and SD across the three fits, not a
  hierarchical variance. The model carries no `eta` terms; the Table 4 section
  propagates the Table 2 SDs by Monte Carlo, as the paper does.
- **Non-positive normal draws are resampled.** The paper assumed the clearances
  are normally distributed but does not say how it handled the negative draws
  that a CV of 32-36 % produces. Resampling is used here; roughly one draw in
  400 is affected, which is well below the two significant figures Table 4
  reports.
- **Equation 9 (MRP4) is not encoded.** The paper simulated two extremes for the
  basolateral efflux transporter, 100 % MRP3 (eq. 8) or 100 % MRP4 (eq. 9),
  found "the simulation results were similar", and selected MRP3 for every
  reported simulation. Eq. 9 is therefore a robustness check rather than a
  reported final model, and is excluded. The Table 1 MRP4 potencies (telmisartan
  11-36 umol/L, bosentan 22 umol/L) are consequently not carried in `ini()`.
- **Ranges in Table 1 are entered as the mean of their endpoints**, following the
  Methods statement that "the mean IC50 values for each transporter in Table 1
  were used". This affects telmisartan BSEP (16-16.2 -> 16.1), bosentan BSEP
  (23-42 -> 32.5) and bosentan NTCP (18 and 36 -> 27). Bosentan OATP1B1 is taken
  as the printed 18 umol/L, per Table 1's footnote a ("Not available and
  therefore assumed to be the same as NTCP").
- **Supplemental Figure 1 was not obtainable.** EuropePMC reports this article as
  not open access and its supplementary-files endpoint returns a gateway error;
  the publisher's supplement URL returns HTTP 403. The figure shows a simulation
  of how telmisartan's effect grows as the uptake phase is extended, generated
  from equations and parameter values that are all in the main text, so no
  parameter is missing. The one claim that depends on it -- "After a 30-minute
  uptake phase, the simulated TCA Ct,Cells for telmisartan based on [I]t,cell was
  threefold of the simulation based on [I]u,cyt" -- is not asserted here; the
  packaged model gives about twofold at 30 minutes for the 10 umol/L level.
- **The observed Cells + Bile point at 20 minutes is under-predicted**, by the
  packaged model and by the paper's own Figure 1B line alike (line ~84 against an
  observed mean ~105 pmol/mg protein). The paper attributes the Figure 1B lines
  to "the mean of best-fit parameter estimates from three SCHH datasets", and a
  mean-of-parameters simulation is not the mean of three simulations, so the two
  need not agree.
