---
title: "Ivermectin (Gwee 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ivermectin (Gwee 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Gwee_2020_ivermectin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Gwee_2020_ivermectin"))$description`
* Article: [PLoS Negl Trop Dis 14(12):e0008886](https://doi.org/10.1371/journal.pntd.0008886)
* Supporting information (S1 Text, S2 Text, S1 Table, S2 Table): <https://journals.plos.org/plosntds/article?id=10.1371/journal.pntd.0008886>

Gwee et al. fitted two models to the same data and published both in the
supporting information:

* `Gwee_2020_ivermectin` -- the authors' **final** model, with the body-weight
  exponents fixed to their allometric "biological prior" values (0.75 on CL/F
  and Q/F, 1 on Vc/F and Vp/F). This is the model the paper uses for its dose
  simulations in children aged 2 to 4 years.
* `Gwee_2020_ivermectin_estimatedExponents` -- the alternative model with the
  weight exponents on CL/F and Vc/F estimated (0.944 and 2.16) and no weight
  effect on Q/F or Vp/F. It had the lower objective function value but, in the
  authors' words, "provided the same inference in terms of dose prediction",
  so it was not carried forward.

# Population

The ITCH (Ivermectin Therapy in Children) study enrolled 26 Indigenous
Australian children with scabies in a remote community in Arnhem Land,
Northern Territory, between April 2016 and March 2018. Eleven (42%) were male;
the median age was 10.9 years (range 5.5 to 14.9) and the median body weight
37.6 kg (18.5 to 74.5). Children received a single oral dose of 200 ug/kg
rounded to the nearest whole or half 3 mg tablet with food (median 190 ug/kg,
range 120 to 230). Each child gave two plasma samples (one child gave one), for
48 concentrations in total, drawn in one of five sampling-design groups that
together span 2 hours to 14 days after the dose (Table 1 of the paper). The
lower limit of quantification was 2 ug/L. Demographics are from the first
paragraph of the Results; the paper has no baseline-characteristics table.

The same information is available programmatically via
`readModelDb("Gwee_2020_ivermectin")()$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Gwee_2020_ivermectin.R` and
`inst/modeldb/specificDrugs/Gwee_2020_ivermectin_estimatedExponents.R`.

| Equation / parameter | Final model | Estimated-exponent model | Source location |
|---|---|---|---|
| Structure: 2 compartments, first-order absorption, linear elimination | -- | -- | Results, paragraph 2 |
| `CL/F = TVCL * (WT/37.55)^e_wt_cl` | 6.94 L/h, exponent 0.75 fixed | 6.73 L/h, exponent 0.944 | S1 / S2 Text; S1 / S2 Table |
| `Vc/F = TVVc * (WT/37.55)^e_wt_vc` | 88 L, exponent 1 fixed | 160 L, exponent 2.16 | S1 / S2 Text; S1 / S2 Table |
| `Q/F` | 13.7 L/h x (WT/37.55)^0.75 (fixed) | 2.95 L/h, no weight effect | S1 / S2 Text; S1 / S2 Table |
| `Vp/F` | 344 L x (WT/37.55) (fixed) | 447 L, no weight effect | S1 / S2 Text; S1 / S2 Table |
| `ka` | 0.5 1/h fixed | 0.5 1/h fixed | S1 / S2 Table |
| BSV %CV (CL, Vc, Q, Vp) | 54, 237.1, 40.8, 38.9 | 68.7, 86.9, 108.7, 29.8 | S1 / S2 Table, read as described below |
| Correlations (CL-Vc, CL-Q, CL-Vp, Vc-Q, Vc-Vp, Q-Vp) | -0.30, 0.70, 0.863, 0.467, 0.215, 0.964 | -0.042, 0.399, -0.958, -0.934, -0.239, -0.125 | S1 / S2 Table, read as described below |
| Residual error, SD of an additive error on the log scale (`Cc ~ lnorm(expSd)`) | 0.198 | 0.062 | S1 / S2 Table |
| `Cc = 1000 * central / vc` (mg and L to ug/L) | -- | -- | Units: dose in mg, concentrations in ug/L (Methods) |

## Reading the variability block of S1 and S2 Tables

Both supporting tables print the between-subject-variability block with the
value column **one row above** its labels. The block has 11 label rows (a
heading, four %CV rows and six correlation rows) but only 10 values: the first
value sits on the heading row, and the last label, "Correlation (Q, Vp)", has
no value. Read literally, the Vp row would carry a %CV of "-30%", which is
impossible. Read in order, the 10 values are the four %CVs (CL, Vc, Q, Vp)
followed by the six correlations in the printed label order, which is how
they are encoded here.

The tables do not say whether "%CV" means `100 * omega` (the standard-deviation
scale) or the exact log-normal CV, `100 * sqrt(exp(omega^2) - 1)`. For the Vc/F
entry of 237.1% the two readings give omega^2 = 5.62 or 1.89, so the choice
matters. The maintainers took the SD-scale reading, `omega = CV/100`, for two
reasons that are both checked below:

1. **Figure 1 (pcVPC).** In the earliest time bin, the model's 10th percentile
   is about 4 ug/L. Under the SD-scale reading the packaged model gives about
   3.3 ug/L; under the log-normal reading it gives about 10.5 ug/L.
2. **Simulated AUC spread.** For one 3 mg tablet in children weighing 10 to 15
   kg the paper reports a median AUC of 976 ug*h/L with an IQR of 671 to 1384.
   The SD-scale reading reproduces the width of that IQR; the log-normal
   reading gives an IQR that is too narrow.

# Virtual cohorts

The observed data are not public. Two virtual populations are used.

* **Study population (children aged 5 to 15 years).** Age is uniform from 5.5
  to 14.9 years. Weight follows an exponential weight-for-age curve with 10.5%
  growth per year and a log-normal SD of 0.2, passing through the reported
  median weight (37.6 kg) at the median age of the virtual cohort (10.2 years),
  so that the virtual median weight matches the study's. Weights outside the observed 18.5 to 74.5 kg range are
  redrawn, not clamped. The dose is 200 ug/kg rounded to the nearest half 3 mg
  tablet (1.5 mg steps).
* **Children aged 2 to 4 years weighing 10 to 15 kg.** The paper used a
  separate covariate database of 288 children that is not published, so weight
  is taken as uniform from 10 to 15 kg. The same weights are dosed two ways, as
  in the paper: (i) 200 ug/kg rounded to the nearest half or whole 3 mg tablet
  (1.5 mg below 11.25 kg, 3 mg above), and (ii) one 3 mg tablet.

```{r cohort}
rxode2::rxSetSeed(20201207)
set.seed(20201207)
n_arm <- 200

draw_study_weights <- function(age) {
  wt <- 37.6 * exp(0.105 * (age - 10.2) + rnorm(length(age), 0, 0.2))
  bad <- wt < 18.5 | wt > 74.5
  while (any(bad)) {
    wt[bad] <- 37.6 * exp(0.105 * (age[bad] - 10.2) + rnorm(sum(bad), 0, 0.2))
    bad <- wt < 18.5 | wt > 74.5
  }
  wt
}

round_to_half_tablet <- function(dose_mg) {
  pmax(1.5, round(dose_mg / 1.5) * 1.5)
}

age_study <- runif(n_arm, 5.5, 14.9)
wt_study <- draw_study_weights(age_study)
wt_young <- runif(n_arm, 10, 15)

subjects <- dplyr::bind_rows(
  tibble(
    id = seq_len(n_arm), treatment = "Study population, 200 ug/kg rounded",
    AGE = age_study, WT = wt_study, amt = round_to_half_tablet(0.2 * wt_study)
  ),
  tibble(
    id = n_arm + seq_len(n_arm), treatment = "Age 2-4 y, 200 ug/kg rounded",
    AGE = NA_real_, WT = wt_young, amt = round_to_half_tablet(0.2 * wt_young)
  ),
  tibble(
    id = 2 * n_arm + seq_len(n_arm), treatment = "Age 2-4 y, 3 mg",
    AGE = NA_real_, WT = wt_young, amt = 3
  )
)

# Log-spaced grid: the Vc/F variability is large, so the absorption peak must
# be resolved for every subject; 2000 h is more than 30 typical terminal
# half-lives.
obs_times <- c(0, exp(seq(log(0.05), log(2000), length.out = 90)))

events <- dplyr::bind_rows(
  subjects |> dplyr::mutate(time = 0, evid = 1L, cmt = "depot"),
  subjects |>
    dplyr::select(-amt) |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(evid = 0L, amt = 0, cmt = "central")
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))

subjects |>
  group_by(treatment) |>
  summarise(
    `Median WT (kg)` = median(WT),
    `WT range (kg)` = sprintf("%.1f-%.1f", min(WT), max(WT)),
    `Median dose (ug/kg)` = median(1000 * amt / WT),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 1, caption = "Virtual cohorts.")
```

# Simulation

```{r simulate}
mod <- readModelDb("Gwee_2020_ivermectin")
sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep = c("treatment", "WT", "AGE"),
  returnType = "data.frame"
)
stopifnot(!anyNA(sim$Cc))
```

# Replicate Figure 1 (prediction-corrected VPC)

A prediction-corrected VPC rescales every observation to the typical
prediction of its bin, which removes the effect of dose and body weight. The
equivalent simulation gives every virtual subject the median weight (37.55 kg)
and the median dose (190 ug/kg x 37.6 kg = 7.14 mg), with between-subject and
residual variability. The points are the model percentiles of Figure 1,
digitised by the maintainers (approximate; the paper's percentiles are over
time bins spanning 0 to 15 h, 15 to 40 h, 40 to 105 h, 105 to 220 h and 220 to
340 h).

```{r figure-1}
fig1_published <- tibble::tribble(
  ~time, ~p10, ~p50, ~p90,
  3.6, 4.3, 35.2, 50.7,
  25, 3.66, 6.3, 14.3,
  49, 1.70, 3.38, 5.46,
  177, 0.177, 1.12, 4.94,
  283, 0.113, 0.31, 2.68
)

vpc_times <- c(0.5, 1, 2, 3, 3.6, 4, 6, 8, 12, 18, 25, 36, 49, 72, 100, 140, 177, 220, 283, 340)
vpc_events <- tibble(id = seq_len(n_arm), WT = 37.55) |>
  tidyr::crossing(time = c(0, vpc_times)) |>
  dplyr::mutate(
    evid = ifelse(time == 0, 1L, 0L),
    amt = ifelse(time == 0, 7.14, 0),
    cmt = ifelse(time == 0, "depot", "central")
  )
vpc_events <- dplyr::bind_rows(
  vpc_events,
  vpc_events |> dplyr::filter(time == 0) |> dplyr::mutate(evid = 0L, amt = 0, cmt = "central")
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

vpc_sim <- rxode2::rxSolve(mod, events = vpc_events, returnType = "data.frame") |>
  dplyr::filter(time > 0)

vpc_q <- vpc_sim |>
  group_by(time) |>
  summarise(
    p10 = quantile(sim, 0.10), p50 = quantile(sim, 0.50), p90 = quantile(sim, 0.90),
    .groups = "drop"
  )

ggplot(vpc_q, aes(time)) +
  geom_ribbon(aes(ymin = p10, ymax = p90), alpha = 0.2, fill = "steelblue") +
  geom_line(aes(y = p50), colour = "steelblue") +
  geom_point(
    data = tidyr::pivot_longer(fig1_published, -time, names_to = "pct", values_to = "conc"),
    aes(time, conc, shape = pct), colour = "red"
  ) +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)", y = "Ivermectin (ug/L)", shape = "Figure 1",
    title = "Figure 1 - 10th, 50th and 90th percentiles",
    caption = "Band and line: packaged model. Red points: digitised from Figure 1 of Gwee 2020."
  )
```

```{r figure-1-table}
fig1_published |>
  dplyr::inner_join(vpc_q, by = "time", suffix = c("_published", "_model")) |>
  dplyr::select(time, p10_published, p10_model, p50_published, p50_model, p90_published, p90_model) |>
  dplyr::rename(
    "Time (h)" = time,
    "P10 Figure 1" = p10_published, "P10 model" = p10_model,
    "P50 Figure 1" = p50_published, "P50 model" = p50_model,
    "P90 Figure 1" = p90_published, "P90 model" = p90_model
  ) |>
  knitr::kable(digits = 2, caption = "Figure 1 percentiles (digitised) vs. the packaged model (ug/L).")

fig1_check <- dplyr::inner_join(fig1_published, vpc_q, by = "time", suffix = c("_pub", "_mod"))
stopifnot(nrow(fig1_check) == 5)
# The median is the structural check: dose, CL/F, Vc/F and the distribution
# phase all set it. Its first-bin value is taken at 3.6 h on a steep rising
# and falling peak, so allow for the bin position.
stopifnot(abs(median(log(fig1_check$p50_mod / fig1_check$p50_pub))) < log(1.3))
```

The model's median tracks the published median from the absorption peak to
the last bin, within about a third at every bin and more closely at most.
The published percentiles are prediction-corrected within wide time bins that
each hold roughly 8 to 14 of the 48 observations, so the 10th and 90th
percentiles agree only in magnitude; the flat published 90th percentile
between 50 and 180 h is a feature of that binning.

## Which %CV scale Figure 1 supports

The earliest bin is where the two readings of the %CV column separate. A
10th percentile estimated from 200 random subjects varies by about a factor of
two between draws here, because the Vc/F variability is so large. The chunk
below therefore uses a fixed, stratified set of 199 random-effect vectors (a
rank-1 lattice mapped through the normal quantile function and the Cholesky
factor of the omega matrix), so the result is the same on every machine. It
solves the typical subject of Figure 1 at 4 h under the packaged omega and
under the omega rebuilt with the log-normal reading and the same correlations.

```{r omega-scale-vpc}
ui <- rxode2::rxode(mod)
omega_sd <- ui$omega
cv <- sqrt(diag(omega_sd))
sd_lognormal <- sqrt(log(1 + cv^2))
omega_lognormal <- diag(sd_lognormal) %*% stats::cov2cor(omega_sd) %*% diag(sd_lognormal)
dimnames(omega_lognormal) <- dimnames(omega_sd)

# Rank-1 lattice: n prime, each coordinate takes every stratum (k + 0.5)/n once.
lattice_etas <- function(omega, n = 199, gen = c(1, 45, 70, 110)) {
  u <- (outer(seq_len(n) - 1, gen) %% n + 0.5) / n
  eta <- stats::qnorm(u) %*% chol(omega)
  colnames(eta) <- colnames(omega)
  as.data.frame(eta)
}

p10_at_4h <- function(omega) {
  etas <- lattice_etas(omega)
  ev <- data.frame(
    id = rep(seq_len(nrow(etas)), each = 2), time = c(0, 4), evid = c(1L, 0L),
    amt = c(7.14, 0), cmt = c("depot", "central"), WT = 37.55
  )
  # The random effects are supplied as parameters, so rxode2's note that a
  # multi-subject solve has no omega is expected here.
  s <- withCallingHandlers(
    rxode2::rxSolve(rxode2::zeroRe(ui), events = ev, params = etas, returnType = "data.frame"),
    warning = function(w) {
      if (grepl("without 'omega'", conditionMessage(w))) invokeRestart("muffleWarning")
    }
  )
  stopifnot(nrow(s) == nrow(etas), !anyNA(s$Cc))
  unname(quantile(s$Cc, 0.10))
}

p10_tab <- tibble(
  Reading = c("Published Figure 1 (first bin)", "SD scale, omega = CV (packaged)", "Log-normal CV"),
  `10th percentile at 4 h (ug/L)` = c(4.3, p10_at_4h(omega_sd), p10_at_4h(omega_lognormal))
)
knitr::kable(p10_tab, digits = 2)

# Deterministic. Measured 3.32 vs 10.54 ug/L; other lattice generators give
# 3.3-3.6 vs 10.5-11.4. The published value is about 4.3.
stopifnot(
  abs(log(p10_tab[[2]][2] / 4.3)) < abs(log(p10_tab[[2]][3] / 4.3)),
  p10_tab[[2]][2] > 2.5, p10_tab[[2]][2] < 5,
  p10_tab[[2]][3] > 8
)
```

# Typical-value AUC against the paper's simulations

For a linear model the median AUC of a cohort equals the typical-subject AUC,
`1000 * Dose / CL`, at the median covariate. This is a deterministic check of
the clearance, its weight exponent and the units.

```{r typical-auc}
typical_auc <- function(model, dose_mg, wt) {
  ev <- tibble(id = 1L, time = c(0, 1), evid = c(1L, 0L), amt = c(dose_mg, 0),
               cmt = c("depot", "central"), WT = wt)
  s <- rxode2::rxSolve(rxode2::zeroRe(model), events = ev, returnType = "data.frame")
  1000 * dose_mg / s$cl[1]
}

typ_tab <- tibble::tribble(
  ~Scenario, ~dose, ~wt, ~published,
  "Study population: 190 ug/kg at 37.6 kg", 0.190 * 37.6, 37.6, 1001,
  "One 3 mg tablet, 10-15 kg (at 12.5 kg)", 3, 12.5, 976,
  "One 3 mg tablet, 12-15 kg (at 13.5 kg)", 3, 13.5, 953
) |>
  dplyr::rowwise() |>
  dplyr::mutate(model = typical_auc(mod, dose, wt)) |>
  dplyr::ungroup() |>
  dplyr::mutate(pct_diff = 100 * (model / published - 1))

typ_tab |>
  dplyr::rename(
    "Dose (mg)" = dose, "WT (kg)" = wt,
    "Published median AUC (ug*h/L)" = published,
    "Typical-value AUC (ug*h/L)" = model, "% diff" = pct_diff
  ) |>
  knitr::kable(digits = 1)

stopifnot(all(abs(typ_tab$pct_diff) < 5))
```

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)
# Numeric-path undershoot in the far tail is integrator noise; floor it.
stopifnot(all(sim_nca$Cc >= -1e-6 * max(sim_nca$Cc)))
sim_nca <- sim_nca |> dplyr::mutate(Cc = pmax(Cc, 0))
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time) |>
  dplyr::group_by(id) |>
  dplyr::filter(time <= time[which.max(Cc)] | Cc >= 1e-6 * max(Cc)) |>
  dplyr::ungroup()

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)
intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

auc_subj <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD == "aucinf.obs") |>
  dplyr::select(id, treatment, aucinf.obs = PPORRES) |>
  dplyr::left_join(
    sim |> dplyr::filter(time == 0) |> dplyr::distinct(id, cl, WT, AGE),
    by = "id"
  ) |>
  dplyr::left_join(dose_df |> dplyr::select(id, amt), by = "id")

# Mass balance: PKNCA AUCinf must equal 1000 * Dose / CL for every subject.
auc_ratio <- auc_subj$aucinf.obs / (1000 * auc_subj$amt / auc_subj$cl)
stopifnot(abs(median(auc_ratio) - 1) < 0.01, quantile(abs(auc_ratio - 1), 0.9) < 0.03)
```

## Comparison against published AUC medians

The paper reports the median AUC for the enrolled children overall and by age
group (from individual clearances), and for the simulated children aged 2 to 4
years by dosing strategy and weight band. The weight bands for the young
children split at 11.25 kg, the weight at which 200 ug/kg rounds up from 1.5 mg
to 3 mg.

```{r nca-compare}
sim_groups <- dplyr::bind_rows(
  auc_subj |> dplyr::filter(treatment == "Study population, 200 ug/kg rounded") |>
    dplyr::mutate(group = "Study population, all"),
  auc_subj |> dplyr::filter(treatment == "Study population, 200 ug/kg rounded", AGE <= 11) |>
    dplyr::mutate(group = "Study population, 5-11 y"),
  auc_subj |> dplyr::filter(treatment == "Study population, 200 ug/kg rounded", AGE > 11) |>
    dplyr::mutate(group = "Study population, >11 y"),
  auc_subj |> dplyr::filter(treatment == "Age 2-4 y, 200 ug/kg rounded") |>
    dplyr::mutate(group = "2-4 y, 200 ug/kg rounded, all"),
  auc_subj |> dplyr::filter(treatment == "Age 2-4 y, 200 ug/kg rounded", WT < 11.25) |>
    dplyr::mutate(group = "2-4 y, 200 ug/kg rounded, <11 kg"),
  auc_subj |> dplyr::filter(treatment == "Age 2-4 y, 3 mg") |>
    dplyr::mutate(group = "2-4 y, 3 mg, all"),
  auc_subj |> dplyr::filter(treatment == "Age 2-4 y, 3 mg", WT < 11.25) |>
    dplyr::mutate(group = "2-4 y, 3 mg, 10-11 kg"),
  auc_subj |> dplyr::filter(treatment == "Age 2-4 y, 3 mg", WT >= 11.25) |>
    dplyr::mutate(group = "2-4 y, 3 mg, 12-15 kg")
) |>
  dplyr::group_by(group) |>
  dplyr::summarise(aucinf.obs = median(aucinf.obs), n = dplyr::n(), .groups = "drop")
stopifnot(nrow(sim_groups) == 8, all(sim_groups$n >= 20))

published_auc <- tibble::tribble(
  ~group, ~aucinf.obs,
  "Study population, all", 1001,
  "Study population, 5-11 y", 895,
  "Study population, >11 y", 1173,
  "2-4 y, 200 ug/kg rounded, all", 917,
  "2-4 y, 200 ug/kg rounded, <11 kg", 620,
  "2-4 y, 3 mg, all", 976,
  "2-4 y, 3 mg, 10-11 kg", 1240,
  "2-4 y, 3 mg, 12-15 kg", 953
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_groups |> dplyr::select(group, aucinf.obs),
  reference = published_auc,
  by = "group",
  units = c(aucinf.obs = "ug*h/L"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Median AUC, simulated vs. published. * differs from reference by >20%.")

pct <- 100 * (sim_groups$aucinf.obs[match(published_auc$group, sim_groups$group)] /
  published_auc$aucinf.obs - 1)
names(pct) <- published_auc$group
# Structural gate on the centre: a mis-transcribed clearance, exponent or unit
# moves every group by tens of percent.
stopifnot(abs(median(pct)) < 10)
# Envelope only on groups of at least 100 virtual subjects (sampling error of
# a median about 7% or less). The enrolled age groups (published medians of
# about 13 children each) and the 10-11 kg bands (about 50 virtual subjects
# each here) scatter by 20% or more between draws and are shown, not gated.
envelope_groups <- setdiff(
  sim_groups$group[sim_groups$n >= 100],
  c("Study population, 5-11 y", "Study population, >11 y")
)
stopifnot(length(envelope_groups) >= 3, all(abs(pct[envelope_groups]) < 25))
```

The overall medians, for the enrolled children and for both dosing strategies
in children aged 2 to 4 years, are reproduced within about 15%. The subgroups
scatter more. Each enrolled age group is the median of about 13 children, and
each 10 to 11 kg band holds only about 50 virtual subjects here, so a single
subgroup can differ from the paper by 20% or more between draws. The paper's
split of 895 vs. 1173 ug*h/L between the age groups is also larger than a model
can produce when AUC rises only as `WT^0.25` at a fixed mg/kg dose. The
deterministic typical-value table above is the tighter check on the weight
bands.

## Which %CV scale the simulated AUC spread supports

With AUC = `1000 * Dose / CL`, the spread of log AUC in a one-tablet cohort is
the spread of the CL/F random effect plus the spread of `0.75 * log(WT)`. The
chunk below computes the IQR exactly (no simulation) for weights uniform on the
paper's band, under both readings of the 54% CV on CL/F.

```{r omega-scale-auc}
log_iqr <- function(omega_cl, wt_lo, wt_hi) {
  wt <- seq(wt_lo, wt_hi, length.out = 401)
  mu <- log(3000 / (6.94 * (wt / 37.55)^0.75))
  cdf <- function(x) mean(stats::pnorm(x, mu, omega_cl))
  q <- vapply(c(0.25, 0.75), function(p) {
    stats::uniroot(function(x) cdf(x) - p, range(mu) + c(-5, 5))$root
  }, numeric(1))
  diff(q)
}
iqr_tab <- tibble::tribble(
  ~Band, ~lo, ~hi, ~published,
  "3 mg, 10-15 kg", 10, 15, log(1384 / 671),
  "3 mg, 12-15 kg", 11.25, 15, log(1357 / 649)
) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    sd_scale = log_iqr(sqrt(omega_sd["etalcl", "etalcl"]), lo, hi),
    lognormal = log_iqr(sqrt(log(1 + omega_sd["etalcl", "etalcl"])), lo, hi)
  ) |>
  dplyr::ungroup()

iqr_tab |>
  dplyr::select(Band, published, sd_scale, lognormal) |>
  dplyr::rename(
    "Published log(Q3/Q1)" = published,
    "SD scale, omega = CV (packaged)" = sd_scale,
    "Log-normal CV" = lognormal
  ) |>
  knitr::kable(digits = 3)

stopifnot(all(abs(iqr_tab$sd_scale - iqr_tab$published) <
  abs(iqr_tab$lognormal - iqr_tab$published)))
```

The SD-scale reading reproduces both published IQR widths to within 0.02 on
the log scale. The log-normal reading is 0.03 to 0.05 too narrow. That
difference is only about one to two sampling SDs of a 1000-subject simulation,
so on its own this is supporting evidence; the Figure 1 comparison above is
the decisive one.

# Estimated-exponent model

The paper states that the estimated-exponent model "provided the same
inference in terms of dose prediction". The typical-value AUCs of the two
models are compared below for the study median and for the young children.

```{r estimated-exponents}
mod_est <- readModelDb("Gwee_2020_ivermectin_estimatedExponents")
cmp_models <- tibble::tribble(
  ~Scenario, ~dose, ~wt,
  "Study median: 7.14 mg at 37.6 kg", 7.14, 37.6,
  "3 mg at 10.5 kg", 3, 10.5,
  "3 mg at 12.5 kg", 3, 12.5,
  "3 mg at 14.5 kg", 3, 14.5
) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    final = typical_auc(mod, dose, wt),
    estimated = typical_auc(mod_est, dose, wt)
  ) |>
  dplyr::ungroup()

cmp_models |>
  dplyr::rename(
    "Dose (mg)" = dose, "WT (kg)" = wt,
    "Final model AUC (ug*h/L)" = final,
    "Estimated-exponent model AUC (ug*h/L)" = estimated
  ) |>
  knitr::kable(digits = c(NA, 2, 1, 0, 0))

# The paper compared the two models in the study population (S2 Fig) and
# found the same inference; at the study median they agree closely.
stopifnot(abs(cmp_models$estimated[1] / cmp_models$final[1] - 1) < 0.05)
# Deterministic claims made in the prose below.
young <- cmp_models$wt < 20
stopifnot(
  all(cmp_models$final[young] > 0.8 * 1001 & cmp_models$final[young] < 1.25 * 1001),
  all((cmp_models$estimated[young] > 1.25 * 1001) == (cmp_models$wt[young] <= 12.5))
)

# Closed-form check of the estimated-exponent clearance (S2 Text).
stopifnot(abs(cmp_models$estimated[1] / (7140 / (6.73 * (37.6 / 37.55)^0.944)) - 1) < 1e-6)

est_sim <- rxode2::rxSolve(
  mod_est, events = events |> dplyr::filter(treatment == "Age 2-4 y, 3 mg"),
  returnType = "data.frame"
)
stopifnot(!anyNA(est_sim$Cc))
```

At the study median the two models agree within about 3%, which is the
comparison the paper made (S2 Fig, study population only). Extrapolated to
children of 10 to 15 kg, the estimated-exponent model predicts about 25 to 30%
higher exposure than the final model, because its CL/F exponent (0.944) is
steeper than 0.75. For one 3 mg tablet, the final model keeps the typical AUC
inside the paper's equivalence window of 80 to 125% of 1001 ug*h/L (801 to 1251)
across 10.5 to 14.5 kg, while the estimated-exponent model puts children of
12.5 kg and below above it. The paper's Methods give extrapolation to younger
children as the reason for preferring the fixed allometric exponents.

# Assumptions and deviations

- **Variability block alignment.** S1 and S2 Tables print the between-subject
  variability values one row above their labels. The values were read in
  order as four %CVs followed by six correlations; see "Reading the
  variability block" above.
- **%CV scale.** The %CVs were taken as `100 * omega` (SD scale), so each
  variance is `(CV/100)^2`. The paper does not define the scale. The
  maintainers chose this reading because it reproduces the early 10th
  percentile of Figure 1 and the published AUC IQRs, and the exact log-normal
  reading does not.
- **Positive-definite nudge.** The printed correlations are rounded, and the
  rounded matrices are slightly non-positive-definite (smallest eigenvalue
  -2e-5 for the final model and -0.001 for the estimated-exponent model). Every
  off-diagonal element was multiplied by 0.99 so that the matrices can be
  sampled. The variances are as printed.
- **Residual error.** "sigma additive in log domain" is encoded as
  `Cc ~ lnorm(expSd)`.
- **Bioavailability.** The paper reports apparent parameters (CL/F, Vc/F, Q/F,
  Vp/F) and no bioavailability term, so the dose enters the depot with F = 1.
- **Virtual cohorts.** The weight-for-age curve for the enrolled children and
  the uniform 10 to 15 kg weights for the children aged 2 to 4 years are
  maintainers' assumptions. The paper's covariate database of 288 children is
  not published.
- **Figure 1 points** were digitised by the maintainers and are approximate.
  The paper's pcVPC uses time bins, so its percentiles are not directly
  comparable to a fixed-time simulation.
- **Screened covariates.** Age and sex were tested and not retained; they are
  recorded in `covariatesDataExcluded`.
- No erratum or correction notice for this article was found (checked
  2026-09-27). The publisher replaced the uncorrected proof of 7 December 2020
  with a final version on 17 December 2020; every value used here is the same
  in both.
