---
title: "Talazoparib + enzalutamide combination therapy (Hadigol 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Talazoparib + enzalutamide combination therapy (Hadigol 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Hadigol_2026_talazoparib_enzalutamide"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Article: <https://doi.org/10.1002/jcph.70125>\n")
cat("* Trial: TALAPRO-2, <https://clinicaltrials.gov/study/NCT03395197>\n")
```

This vignette validates a single model file that carries **three coupled
analytes**: talazoparib, enzalutamide, and the active metabolite N-desmethyl
enzalutamide. The three cannot be separated into independent models, because
talazoparib's apparent clearance is a function of the instantaneous plasma
concentration of the other two -- simulating talazoparib requires simulating the
enzalutamide sub-system alongside it.

# Population

The model was fit to 811 men with metastatic castration-resistant prostate
cancer (mCRPC), pooled from Part 1 (19 patients, open-label, non-randomized,
dense sampling) and Part 2 Cohort 1 (792 patients, randomized, double-blind,
sparse sampling) of the Phase 3 TALAPRO-2 trial. Part 2 Cohort 1 enrolled
patients **unselected** for homologous recombination repair (HRR) gene
deficiencies; the HRR-deficient Cohort 2 was not pooled into this analysis.

414 patients were randomized to talazoparib plus enzalutamide and 397 to
placebo plus enzalutamide, so all 811 contribute enzalutamide and N-desmethyl
enzalutamide data while only the 412 of 414 with at least one non-BLQ
talazoparib observation contribute talazoparib data. The dataset comprised 6030
plasma observations for enzalutamide and for N-desmethyl enzalutamide, and 2961
for talazoparib.

Enzalutamide was dosed at 160 mg once daily with protocol-permitted reductions
to 120 mg and 80 mg. Talazoparib was dosed at 0.5 mg once daily, reduced to a
0.35 mg once-daily starting dose in moderate renal impairment. Part 1 initially
combined talazoparib 1 mg once daily with enzalutamide, which produced roughly
two-fold higher than expected talazoparib troughs at Week 5 along with more
Grade 3/4 haematologic adverse events than expected; that is the observation the
drug-drug interaction term in this model quantifies, and the reason the
combination dose is 0.5 mg rather than 1 mg.

The medians named in the main text are baseline body weight 80.1 kg and age 71
years (both stated as population medians), with 86.85 mL/min the baseline
creatinine clearance of the typical patient used as the reference in the
covariate equation. Tables S1 and S2 (study-population summary and baseline
demographics) are **not open access**, so covariate *ranges* and the
race/ethnicity and regional breakdown cannot be recorded here -- see
"Assumptions and deviations".

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Hadigol_2026_talazoparib_enzalutamide")()$population`).

# Model structure

Hadigol 2026 Figure 1 gives the enzalutamide parent-metabolite schematic; the
talazoparib arm is described in the Drug-Drug Interactions section. The packaged
model reproduces all three:

**Enzalutamide** (suffix `_enz`) is absorbed first-order from `depot_enz` into
`central_enz`, distributes to `peripheral1_enz`, and leaves `central_enz` at the
full apparent clearance `CLe/Fe`. A fraction `Fmet` of that flux is routed into
the metabolite; the remaining `(1 - Fmet)` is true elimination. Writing it this
way keeps total enzalutamide apparent clearance equal to `CLe/Fe` for **any**
value of `Fmet` -- which is exactly why `Fmet` and the metabolite volume `Vcn`
are not simultaneously identifiable from parent and metabolite plasma data
alone. Hadigol 2026 resolved this by **fixing** `Fmet` to 0.634 from a published
enzalutamide PBPK model, having found that fit better than the alternative
literature device of assuming equal central volumes for parent and metabolite.

**N-desmethyl enzalutamide** (suffix `_ndmenz`) is two-compartment with no
depot: it is formed directly into plasma.

**Talazoparib** carries the bare canonical names, because it is the substrate of
interest. It is two-compartment with first-order absorption and -- unlike the
talazoparib monotherapy model -- no absorption lag time. Its apparent clearance
is built in two stages: equation (6) applies the renal-function power effect to
give the interaction-free base clearance `CLt0/Ft`, then equation (1) applies
the P-glycoprotein inhibition term

$$\mathrm{CL}_t/F_t = \mathrm{CL}_{t0}/F_t \times \bigl(1 - \theta_{slope} \times (C_e + C_n)\bigr)$$

so the clearance is **time-varying**, tracking the perpetrator concentrations
through the dosing interval. The two perpetrator species enter with equal
weight because Hadigol 2026 considered them equipotent on the basis of in vitro
data. When the enzalutamide depot is never dosed, both perpetrator compartments
stay at zero, the bracket evaluates to exactly 1, and talazoparib clearance
falls back to its uninhibited base value.

The three sub-models were fit **sequentially** (enzalutamide; then the
metabolite with enzalutamide parameters fixed to empirical Bayes estimates; then
talazoparib with both fixed), because a simultaneous parent-metabolite fit was
not stable. There is therefore no cross-analyte random-effect covariance to
carry, and the packaged model has three separate variability blocks.

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Hadigol_2026_talazoparib_enzalutamide.R` carries an
in-file comment naming its source row, together with the reported SE, RSE and
the PsN sampling-importance-resampling median and 95% CI. The table below
collects them.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Analyte, ~Parameter, ~Value, ~Source,
  "Talazoparib", "`lcl` (CLt0/Ft)",            "5.078 L/h",        "Table 3",
  "Talazoparib", "`lvc` (Vct/Ft)",             "14.389 L",         "Table 3",
  "Talazoparib", "`lvp` (Vpt/Ft)",             "382.135 L",        "Table 3",
  "Talazoparib", "`lq` (Qt/Ft)",               "15.568 L/h",       "Table 3",
  "Talazoparib", "`lka` (kat)",                "0.157 1/h",        "Table 3",
  "Talazoparib", "`lfdepot` (Ft)",             "1 (fixed)",        "Table 3",
  "Talazoparib", "`e_crcl_cl`",                "0.455",            "Table 3 + equation (6)",
  "Talazoparib", "`e_cenz_cl` (theta slope)",  "6.58e-6 mL/ng",    "Table 3 + equation (1)",
  "Talazoparib", "`etalcl`, `etalvc`",         "0.073, 2.582",     "Table 3 (variances)",
  "Talazoparib", "`expSd`",                    "0.354",            "Table 3 'Thetarized sigma'",
  "Enzalutamide", "`lcl_enz` (CLe/Fe)",        "0.425 L/h",        "Table 1",
  "Enzalutamide", "`lvc_enz` (Vce/Fe)",        "25.293 L",         "Table 1",
  "Enzalutamide", "`lvp_enz` (Vpe/Fe)",        "45.928 L",         "Table 1",
  "Enzalutamide", "`lq_enz` (Qe/Fe)",          "20.644 L/h",       "Table 1",
  "Enzalutamide", "`lka_enz` (kae)",           "3.431 1/h",        "Table 1",
  "Enzalutamide", "`lfdepot_enz` (Fe)",        "1 (fixed)",        "Table 1",
  "Enzalutamide", "`e_wt_cl_enz`",             "0.549 (power)",    "Table 1 + equation (2)",
  "Enzalutamide", "`e_age_cl_enz`",            "-0.003 (linear)",  "Table 1 + equation (2)",
  "Enzalutamide", "`e_wt_vc_enz`",             "3.495 (power)",    "Table 1 + equation (3)",
  "Enzalutamide", "`e_age_vc_enz`",            "1.223 (power)",    "Table 1 + equation (3)",
  "Enzalutamide", "`etalcl_enz`+`etalvc_enz`", "0.037, -0.015, 0.350", "Table 1 block Omega",
  "Enzalutamide", "`expSd_enz`",               "0.145",            "Table 1 'Thetarized sigma'",
  "N-desmethyl", "`lcl_ndmenz` (CLn)",         "0.286 L/h",        "Table 2",
  "N-desmethyl", "`lvc_ndmenz` (Vcn)",         "44.944 L",         "Table 2",
  "N-desmethyl", "`lvp_ndmenz` (Vpn)",         "52.373 L",         "Table 2",
  "N-desmethyl", "`lq_ndmenz` (Qn)",           "9.443 L/h",        "Table 2",
  "N-desmethyl", "`fm` (Fmet)",                "0.634 (fixed)",    "Table 2; fixed from a published enzalutamide PBPK model",
  "N-desmethyl", "`e_wt_cl_ndmenz`",           "0.006 (linear)",   "Table 2 + equation (4)",
  "N-desmethyl", "`e_wt_vc_ndmenz`",           "0.027 (linear)",   "Table 2 + equation (5)",
  "N-desmethyl", "`etalcl_ndmenz`+`etalvc_ndmenz`", "0.057, 0.006, 0.342", "Table 2 block Omega",
  "N-desmethyl", "`expSd_ndmenz`",             "0.125",            "Table 2 'Thetarized sigma'"
) |>
  knitr::kable(caption = "Source trace. Table numbers are Hadigol 2026 main-text tables; equation numbers are its numbered display equations.")
```

# Typical-value simulation machinery

Every published number this vignette reproduces is a **typical-value**
prediction (Hadigol 2026: "Simulations were performed at typical values of PK
parameters"), so the helper below suppresses all random effects with
`omega = NA` / `sigma = NA`. That makes the checks in the next three sections
deterministic: the only difference between the model and the paper is numerical
(observation-grid resolution), not a draw. Tight bounds are therefore correct
here, and are used.

`useLinCmt = FALSE` is required: this is a multi-output model, and rxode2's
automatic ODE-to-`linCmt()` conversion corrupts the endpoint mapping for such
models. Observation rows point at the ODE state `central` and nominate the
endpoint with `dvid = 1L`; rxode2 returns all three algebraic concentrations as
columns regardless.

```{r typical-helper}
mod <- readModelDb("Hadigol_2026_talazoparib_enzalutamide")

# Long enough for the slowest analyte to reach steady state: N-desmethyl
# enzalutamide has a terminal half-life near 10 days, so 150 daily doses is
# about 15 terminal half-lives.
n_day <- 150L
tau <- 24

#' Solve one typical-value regimen and return the final dosing interval.
solve_typical <- function(tala_dose, enza_dose, crcl, wt = 80.1, age = 71,
                          grid = 0.1, params = NULL) {
  t_ss <- (n_day - 1L) * tau
  dose_rows <- data.frame(
    time = 0,
    amt = c(tala_dose, enza_dose),
    evid = 1L,
    cmt = c("depot", "depot_enz"),
    dvid = NA_integer_,
    ii = tau,
    addl = n_day - 1L
  )
  obs_rows <- data.frame(
    time = seq(t_ss, t_ss + tau, by = grid),
    amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L, ii = 0, addl = 0L
  )
  ev <- dplyr::bind_rows(dose_rows, obs_rows)
  ev$id <- 1L
  ev$WT <- wt
  ev$AGE <- age
  ev$CRCL <- crcl

  rxode2::rxSolve(
    mod, ev,
    omega = NA, sigma = NA, params = params,
    useLinCmt = FALSE, returnType = "data.frame", addDosing = FALSE
  ) |>
    dplyr::filter(.data$time >= t_ss) |>
    dplyr::mutate(tad = .data$time - t_ss)
}
```

## Equations (2) to (5): typical covariate-model values

At the reference covariate values (80.1 kg, 71 years) every covariate term must
collapse to 1, so the four covariate-carrying parameters must equal their
printed typical values exactly. This is the cheapest possible check that the
covariate equations were transcribed with the right centring, the right
reference values and the right functional forms (power versus linear).

```{r eq2to5}
ref_solve <- solve_typical(0.5, 160, crcl = 90)

eq_chk <- tibble::tribble(
  ~Parameter,        ~Published, ~Model,
  "CLe/Fe (L/h)",     0.425,      ref_solve$cl_enz[1],
  "Vce/Fe (L)",      25.293,      ref_solve$vc_enz[1],
  "CLn (L/h)",        0.286,      ref_solve$cl_ndmenz[1],
  "Vcn (L)",         44.944,      ref_solve$vc_ndmenz[1]
) |>
  dplyr::mutate(`Rel. error` = abs(.data$Model / .data$Published - 1))

# Pure algebra at the reference covariate values -- must be exact to machine
# precision. A wrong reference value, a power term written as linear (or vice
# versa), or a sign error on the age coefficient all break this immediately.
stopifnot(max(eq_chk$`Rel. error`) < 1e-10)

eq_chk |>
  knitr::kable(digits = c(0, 3, 6, 12),
               caption = "Equations (2)-(5) at the reference covariates (80.1 kg, 71 years).")
```

## Table 4: renal function on talazoparib base clearance

Hadigol 2026 Table 4 tabulates `CLt0/Ft` at the renal-disease cutoff midpoints
and the percent reduction relative to a baseline creatinine clearance of
90 mL/min. Equation (6) is pure algebra, so this too should match the printed
three-significant-figure values exactly.

```{r table4}
t4 <- tibble::tribble(
  ~Stage,       ~CRCL,  ~`Published CLt0/Ft`, ~`Published % reduction`,
  "1-Normal",    90.0,   5.16,                 0.00,
  "2-Mild",      75.0,   4.75,                 7.95,
  "3-Moderate",  45.0,   3.76,                27.13,
  "4-Severe",    22.5,   2.75,                46.71
) |>
  dplyr::mutate(
    `Model CLt0/Ft` = vapply(.data$CRCL, function(cc) solve_typical(0.5, 160, cc)$cl0[1], numeric(1)),
    `Model % reduction` = 100 * (1 - .data$`Model CLt0/Ft` / .data$`Model CLt0/Ft`[1]),
    `CLt0/Ft diff (%)` = 100 * (.data$`Model CLt0/Ft` / .data$`Published CLt0/Ft` - 1)
  )

# Closed-form check against values the paper printed to 3 significant figures,
# so rounding alone allows about 0.1%. Realised max 0.13%, which is that
# rounding. 1% still goes red on any mis-transcription of the 5.078 intercept,
# the 0.455 exponent or the 86.85 reference.
stopifnot(max(abs(t4$`CLt0/Ft diff (%)`)) < 1)
stopifnot(max(abs(t4$`Model % reduction` - t4$`Published % reduction`)) < 0.5)

t4 |>
  dplyr::rename("Baseline CrCl (mL/min)" = CRCL) |>
  knitr::kable(digits = 2,
               caption = "Replicates Table 4 of Hadigol 2026: effect of baseline creatinine clearance on talazoparib apparent base clearance.")
```

## Structural gates: is the interaction actually wired in?

Two checks that a reader cannot get from any table. Both are deterministic.

The first re-solves an identical regimen with `e_cenz_cl` driven to zero and
asserts the talazoparib profile moves. rxode2 recognises a `cl` / `vc` pair and
can solve a compartmental model analytically from those two variables, silently
discarding the explicit ODEs; if that happened here, or if the interaction term
had been attached to a variable that never reaches `cl`, the two solves would be
identical and every number above would still look plausible.

The second is a steady-state mass balance. Over one dosing interval at steady
state the eliminated amount must equal the dose, which for a time-varying
clearance means the integral of `cl(t) * Cc(t)`. The flux is integrated with
`PKNCA::pk.calc.auc.last()` rather than an inline trapezoid.

```{r structural-gates}
with_ddi <- solve_typical(0.5, 160, crcl = 90)
no_ddi   <- solve_typical(0.5, 160, crcl = 90, params = c(e_cenz_cl = 0))

ddi_shift <- max(abs(no_ddi$Cc - with_ddi$Cc) / pmax(with_ddi$Cc, 1e-12))

# The DDI must move the profile. Realised 0.24; the bound only has to exclude
# "the interaction is dead", so 0.01 is the right threshold.
stopifnot(ddi_shift > 0.01)

# Steady-state mass balance. Cc is ng/mL and cl is L/h, so cl * Cc is ug/h;
# 1000 * integral gives ng, comparable with the dose in ng.
elim_ng <- 1000 * PKNCA::pk.calc.auc.last(
  conc = with_ddi$cl * with_ddi$Cc, time = with_ddi$tad
)
dose_ng <- 0.5 * 1e6

# Exact conservation law, not a fit. Realised ratio 1.0000. This is the gate
# that goes red if rxode2 ever auto-solves the model from cl/vc and discards
# the ODEs, or if the fixed Ft were mis-encoded.
stopifnot(abs(elim_ng / dose_ng - 1) < 0.01)

tibble::tribble(
  ~Check, ~Value, ~Requirement,
  "Max relative change in Cc when theta_slope is set to 0", sprintf("%.3f", ddi_shift), "> 0.01",
  "Integral(cl * Cc) over one SS interval / dose",          sprintf("%.4f", elim_ng / dose_ng), "1 +/- 0.01",
  "Uninhibited AUC0-24,ss (ng*h/mL)", sprintf("%.2f", PKNCA::pk.calc.auc.last(no_ddi$Cc, no_ddi$tad)), "reference",
  "Inhibited AUC0-24,ss (ng*h/mL)",   sprintf("%.2f", PKNCA::pk.calc.auc.last(with_ddi$Cc, with_ddi$tad)), "higher"
) |>
  knitr::kable(caption = "Structural gates. Neither is a published number; both fail loudly if the coupling is broken.")
```

The interaction raises talazoparib steady-state exposure by about 25% at the
160 mg enzalutamide dose, equivalently a ~20% reduction in apparent clearance --
consistent with Hadigol 2026's account of why the combination dose was halved
relative to monotherapy.

# Table 5: effect of enzalutamide dose reduction on talazoparib exposure

This is the paper's headline simulation and the strongest available validation:
six regimens crossing two talazoparib doses / renal strata with three
enzalutamide doses, each reported as steady-state AUC0-24, Cmin and Cmax. NCA is
computed with PKNCA over the final dosing interval.

```{r table5-sim}
regimens <- tibble::tribble(
  ~treatment,                 ~tala, ~enza, ~crcl,
  "0.50 mg tala / 160 mg enza", 0.50,  160,   90,
  "0.50 mg tala / 120 mg enza", 0.50,  120,   90,
  "0.50 mg tala / 80 mg enza",  0.50,   80,   90,
  "0.35 mg tala / 160 mg enza", 0.35,  160,   45,
  "0.35 mg tala / 120 mg enza", 0.35,  120,   45,
  "0.35 mg tala / 80 mg enza",  0.35,   80,   45
) |>
  dplyr::mutate(id = dplyr::row_number())

sim_t5 <- dplyr::bind_rows(lapply(seq_len(nrow(regimens)), function(i) {
  solve_typical(regimens$tala[i], regimens$enza[i], regimens$crcl[i]) |>
    dplyr::mutate(treatment = regimens$treatment[i], id = regimens$id[i])
}))

stopifnot(nrow(sim_t5) > 0, !anyNA(sim_t5$Cc), all(sim_t5$Cc >= 0))
```

```{r table5-nca}
# Time is re-based so the final dosing interval runs 0 to 24 h; at steady state
# AUC over that window IS AUC0-tau,ss. Filter on !is.na(Cc) only -- adding
# `time > 0` or `Cc > 0` would drop the time-zero record PKNCA needs to anchor
# the interval (and here that record is the trough, a real observation).
nca_conc <- sim_t5 |>
  dplyr::filter(!is.na(.data$Cc)) |>
  dplyr::select(id, treatment, time = tad, Cc)

nca_dose <- regimens |>
  dplyr::transmute(id, treatment, time = 0, amt = .data$tala)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(nca_conc), Cc ~ time | treatment + id,
                             concu = "ng/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(as.data.frame(nca_dose), amt ~ time | treatment + id,
                             route = "extravascular", doseu = "mg")

intervals <- data.frame(
  start = 0, end = tau,
  cmax = TRUE, cmin = TRUE, tmax = TRUE, auclast = TRUE, cav = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r table5-compare}
published_t5 <- tibble::tribble(
  ~treatment,                 ~auclast, ~cmin, ~cmax,
  "0.50 mg tala / 160 mg enza", 120.80,  3.91,  6.93,
  "0.50 mg tala / 120 mg enza", 113.79,  3.64,  6.62,
  "0.50 mg tala / 80 mg enza",  107.53,  3.39,  6.34,
  "0.35 mg tala / 160 mg enza", 115.87,  3.99,  6.21,
  "0.35 mg tala / 120 mg enza", 109.15,  3.72,  5.92,
  "0.35 mg tala / 80 mg enza",  103.16,  3.48,  5.66
)

cmp_t5 <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published_t5,
  by = "treatment",
  params = c("auclast", "cmin", "cmax"),
  units = c(auclast = "ng*h/mL", cmin = "ng/mL", cmax = "ng/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp_t5,
  caption = "Replicates Table 5 of Hadigol 2026: simulated vs published steady-state talazoparib exposure. * marks a >20% difference from the reference."
)
```

```{r table5-gate}
# Explicit numeric gate on the same comparison, because ncaComparisonTable
# only stars differences above its tolerance and a silently empty join would
# otherwise render a clean-looking table that tested nothing.
nca_wide <- as.data.frame(nca_res$result) |>
  dplyr::filter(.data$PPTESTCD %in% c("auclast", "cmin", "cmax")) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = "PPTESTCD", values_from = "PPORRES")

gate_t5 <- published_t5 |>
  dplyr::inner_join(nca_wide, by = "treatment", suffix = c("_pub", "_mod")) |>
  dplyr::mutate(
    auc_pct  = 100 * (.data$auclast_mod / .data$auclast_pub - 1),
    cmin_pct = 100 * (.data$cmin_mod / .data$cmin_pub - 1),
    cmax_pct = 100 * (.data$cmax_mod / .data$cmax_pub - 1)
  )

# Guard the join first: an empty or short join would make every all() below
# pass vacuously (pattern 10 of the known-failure catalogue).
stopifnot(nrow(gate_t5) == nrow(published_t5))

# Deterministic typical-value comparison -- both sides evaluate the same
# equations, so the residual is grid resolution only, not a per-subject
# physical mechanism. Realised +0.145% to +0.512% across all 18 cells at a
# 0.1 h grid, all in the same direction (the grid slightly over-resolves the
# peak). 2% keeps ~4x headroom while still going red on any mis-transcribed
# clearance, volume, dose or unit, which move these by tens of percent.
stopifnot(
  max(abs(gate_t5$auc_pct)) < 2,
  max(abs(gate_t5$cmin_pct)) < 2,
  max(abs(gate_t5$cmax_pct)) < 2
)

gate_t5 |>
  dplyr::select(treatment, auc_pct, cmin_pct, cmax_pct) |>
  dplyr::rename(
    "Regimen" = treatment,
    "AUC0-24,ss diff (%)" = auc_pct,
    "Cmin,ss diff (%)" = cmin_pct,
    "Cmax,ss diff (%)" = cmax_pct
  ) |>
  knitr::kable(digits = 3,
               caption = "Percent difference from Hadigol 2026 Table 5, all 18 cells.")
```

The paper's conclusion that reducing enzalutamide from 160 mg to 120 mg or 80 mg
does not warrant a talazoparib dose modification is reproduced directly: the
model gives steady-state AUC0-24 reductions of 5.9% and 11.2% respectively at
0.5 mg talazoparib, against the paper's 5.8% and 10.99%.

```{r table5-figure}
sim_t5 |>
  ggplot(aes(.data$tad, .data$Cc, colour = .data$treatment)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time after dose at steady state (h)",
       y = "Talazoparib concentration (ng/mL)",
       colour = NULL,
       title = "Steady-state talazoparib profiles by combination regimen",
       caption = "Typical-value predictions underlying the Table 5 replication (Hadigol 2026).") +
  theme(legend.position = "bottom")
```

# Perpetrator exposure and the time course of inhibition

The interaction is driven by `Ce + Cn`, which the paper never tabulates. Both
species are long-lived relative to the 24 h dosing interval (enzalutamide
terminal half-life about 5 days, the metabolite about 8 to 10 days), so their
concentrations are nearly flat across the interval and talazoparib clearance is
almost -- but not exactly -- constant.

```{r perpetrator}
perp <- ref_solve |>
  dplyr::summarise(
    `Mean Cc_enz (ng/mL)` = mean(.data$Cc_enz),
    `Mean Cc_ndmenz (ng/mL)` = mean(.data$Cc_ndmenz),
    `Mean Ce + Cn (ng/mL)` = mean(.data$Cc_enz + .data$Cc_ndmenz),
    `Mean CL inhibition factor` = mean(.data$cl) / .data$cl0[1],
    `Peak-to-trough range of Ce + Cn (%)` =
      100 * (max(.data$Cc_enz + .data$Cc_ndmenz) / min(.data$Cc_enz + .data$Cc_ndmenz) - 1)
  )

# Independent sanity check on the perpetrator scale. Dose / (CL * tau) gives
# the average steady-state concentration in closed form, including the fixed
# Fmet for the metabolite. These are algebraic identities for a linear system,
# so a few percent tolerance covers only the within-interval curvature.
ce_closed <- 1000 * 160 / (0.425 * tau)
cn_closed <- 1000 * 0.634 * 160 / (0.286 * tau)
stopifnot(
  abs(perp$`Mean Cc_enz (ng/mL)` / ce_closed - 1) < 0.05,
  abs(perp$`Mean Cc_ndmenz (ng/mL)` / cn_closed - 1) < 0.05
)

# The linear inhibition form is not bounded below, so a large enough
# perpetrator exposure would drive the factor negative. At the typical
# regimen it must sit comfortably inside (0, 1).
stopifnot(perp$`Mean CL inhibition factor` > 0.5,
          perp$`Mean CL inhibition factor` < 1)

perp |>
  tidyr::pivot_longer(dplyr::everything(), names_to = "Quantity", values_to = "Value") |>
  knitr::kable(digits = 3,
               caption = "Perpetrator exposure at enzalutamide 160 mg QD steady state and the resulting talazoparib clearance inhibition (typical values).")
```

```{r perpetrator-figure}
ref_solve |>
  dplyr::select(tad, Enzalutamide = Cc_enz, `N-desmethyl enzalutamide` = Cc_ndmenz) |>
  tidyr::pivot_longer(-tad, names_to = "Analyte", values_to = "Conc") |>
  ggplot(aes(.data$tad, .data$Conc, colour = .data$Analyte)) +
  geom_line(linewidth = 0.7) +
  expand_limits(y = 0) +
  labs(x = "Time after dose at steady state (h)", y = "Concentration (ng/mL)",
       colour = NULL,
       title = "Perpetrator concentrations over one steady-state interval",
       caption = "Enzalutamide 160 mg QD; typical values (Hadigol 2026).") +
  theme(legend.position = "bottom")
```

# Virtual cohort

Original observed data are not publicly available, and the demographics table is
not open access, so the cohort below uses covariate distributions consistent
with the medians the main text names and with the renal-function strata the
paper simulates. It exists to exercise the three variability blocks and to show
the spread of the interaction across subjects -- not to reproduce a published
VPC, since Figure 3's underlying data are not available.

```{r cohort}
# set.seed() seeds R's RNG, which is what draws the covariates below. It does
# NOT seed rxode2's simulation RNG, and rxode2 partitions its streams per
# solver thread, so the eta draws differ on a machine with a different thread
# count. Every assertion below is therefore written to hold for any cohort the
# model can produce (pattern 12 of the known-failure catalogue).
set.seed(20260912)

n_per_arm <- 100L   # cap is 200 per arm; 100 is ample here
n_day_cohort <- 84L # 12 weeks, past steady state for all three analytes

make_arm <- function(n, enza_dose, label, id_offset) {
  subj <- tibble::tibble(
    id = id_offset + seq_len(n),
    WT = pmin(150, pmax(45, rnorm(n, 80.1, 14))),
    AGE = pmin(92, pmax(40, rnorm(n, 71, 8))),
    CRCL = pmin(160, pmax(20, rnorm(n, 86.85, 25))),
    treatment = label
  )
  t_ss <- (n_day_cohort - 1L) * tau
  obs_t <- sort(unique(c(seq(0, t_ss, by = tau), seq(t_ss, t_ss + tau, by = 1))))
  doses <- subj |>
    tidyr::expand_grid(tibble::tibble(
      amt = c(0.5, enza_dose), cmt = c("depot", "depot_enz")
    )) |>
    dplyr::mutate(time = 0, evid = 1L, dvid = NA_integer_,
                  ii = tau, addl = n_day_cohort - 1L)
  obs <- subj |>
    tidyr::expand_grid(tibble::tibble(time = obs_t)) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L,
                  ii = 0, addl = 0L)
  dplyr::bind_rows(doses, obs) |> dplyr::arrange(.data$id, .data$time, dplyr::desc(.data$evid))
}

events <- dplyr::bind_rows(
  make_arm(n_per_arm, 160, "160 mg enzalutamide", id_offset = 0L),
  make_arm(n_per_arm,  80,  "80 mg enzalutamide", id_offset = n_per_arm)
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

```{r cohort-sim}
sim <- rxode2::rxSolve(
  mod, events,
  useLinCmt = FALSE,
  keep = c("treatment", "WT", "AGE", "CRCL"),
  returnType = "data.frame"
)

stopifnot(nrow(sim) > 0, !anyNA(sim$Cc), all(sim$Cc >= 0))

# The linear inhibition term has no lower bound by construction, so confirm
# that no simulated subject is driven to a non-positive clearance. This is a
# structural guard on the extrapolation range, not a published number.
stopifnot(all(sim$cl > 0), all(sim$cl / sim$cl0 <= 1))
```

```{r cohort-figure}
t_ss_c <- (n_day_cohort - 1L) * tau

sim |>
  dplyr::filter(.data$time >= t_ss_c) |>
  dplyr::mutate(tad = .data$time - t_ss_c) |>
  dplyr::group_by(.data$treatment, .data$tad) |>
  dplyr::summarise(
    Q05 = quantile(.data$Cc, 0.05), Q50 = median(.data$Cc),
    Q95 = quantile(.data$Cc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(.data$tad, .data$Q50)) +
  geom_ribbon(aes(ymin = .data$Q05, ymax = .data$Q95), alpha = 0.25) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~treatment) +
  labs(x = "Time after dose at steady state (h)",
       y = "Talazoparib concentration (ng/mL)",
       title = "Simulated steady-state talazoparib exposure, median and 5th-95th percentile",
       caption = "Talazoparib 0.5 mg QD; 100 subjects per enzalutamide arm (Hadigol 2026 model).")
```

```{r cohort-inhibition}
inhib <- sim |>
  dplyr::filter(.data$time >= t_ss_c) |>
  dplyr::group_by(.data$treatment, .data$id) |>
  dplyr::summarise(factor = mean(.data$cl) / dplyr::first(.data$cl0), .groups = "drop") |>
  dplyr::group_by(.data$treatment) |>
  dplyr::summarise(
    `Median CL inhibition factor` = median(.data$factor),
    `5th percentile` = quantile(.data$factor, 0.05),
    `95th percentile` = quantile(.data$factor, 0.95),
    .groups = "drop"
  )

# Cohort-derived, so bounds are on magnitude with generous headroom rather
# than on ordering or on one observed run. The 160 mg arm must show more
# inhibition than the 80 mg arm as an ABSOLUTE statement (the typical-value
# factors are ~0.80 and ~0.90, a 10-point separation that no plausible cohort
# draw closes), and neither arm may leave (0, 1).
stopifnot(
  all(inhib$`Median CL inhibition factor` > 0.5),
  all(inhib$`Median CL inhibition factor` < 1),
  inhib$`Median CL inhibition factor`[inhib$treatment == "160 mg enzalutamide"] < 0.87,
  inhib$`Median CL inhibition factor`[inhib$treatment == "80 mg enzalutamide"] > 0.84
)

inhib |>
  dplyr::rename("Enzalutamide arm" = treatment) |>
  knitr::kable(digits = 3,
               caption = "Distribution of the talazoparib clearance inhibition factor across the simulated cohort.")
```

# PKNCA on the cohort, all three analytes

One PKNCA block per output, each with its own concentration object, over the
final steady-state dosing interval. The metabolite has no dose record of its
own, so its PKNCA dose object carries the enzalutamide dose that ultimately
generates it; only interval-bounded parameters (Cmax, Cmin, AUC over the
interval, Cav) are requested, none of which depends on the dose amount.

```{r cohort-pknca}
enza_by_arm <- c("160 mg enzalutamide" = 160, "80 mg enzalutamide" = 80)

nca_one <- function(conc_col, dose_amt_fn) {
  cdat <- sim |>
    dplyr::filter(.data$time >= t_ss_c) |>
    dplyr::mutate(tad = .data$time - t_ss_c, Cc = .data[[conc_col]]) |>
    dplyr::filter(!is.na(.data$Cc)) |>
    dplyr::select(id, treatment, time = tad, Cc)
  stopifnot(nrow(cdat) > 0)

  ddat <- cdat |>
    dplyr::distinct(.data$id, .data$treatment) |>
    dplyr::mutate(time = 0, amt = dose_amt_fn(.data$treatment))

  co <- PKNCA::PKNCAconc(as.data.frame(cdat), Cc ~ time | treatment + id,
                         concu = "ng/mL", timeu = "h")
  do <- PKNCA::PKNCAdose(as.data.frame(ddat), amt ~ time | treatment + id,
                         route = "extravascular", doseu = "mg")
  PKNCA::pk.nca(PKNCA::PKNCAdata(co, do, intervals = data.frame(
    start = 0, end = tau, cmax = TRUE, cmin = TRUE, tmax = TRUE,
    auclast = TRUE, cav = TRUE
  )))
}

nca_tala   <- nca_one("Cc",        function(tr) 0.5)
nca_enz    <- nca_one("Cc_enz",    function(tr) unname(enza_by_arm[as.character(tr)]))
nca_ndmenz <- nca_one("Cc_ndmenz", function(tr) unname(enza_by_arm[as.character(tr)]))

summarise_nca <- function(res, analyte) {
  as.data.frame(res$result) |>
    dplyr::filter(.data$PPTESTCD %in% c("cmax", "cmin", "tmax", "auclast", "cav")) |>
    dplyr::group_by(.data$treatment, .data$PPTESTCD) |>
    dplyr::summarise(median = median(.data$PPORRES), .groups = "drop") |>
    tidyr::pivot_wider(names_from = "PPTESTCD", values_from = "median") |>
    dplyr::mutate(Analyte = analyte, .before = 1)
}

nca_all <- dplyr::bind_rows(
  summarise_nca(nca_tala, "Talazoparib"),
  summarise_nca(nca_enz, "Enzalutamide"),
  summarise_nca(nca_ndmenz, "N-desmethyl enzalutamide")
)

stopifnot(nrow(nca_all) == 6L, !anyNA(nca_all$auclast))

nca_all |>
  dplyr::rename(
    "Arm" = treatment,
    "Cmax,ss (ng/mL)" = cmax,
    "Cmin,ss (ng/mL)" = cmin,
    "Tmax (h)" = tmax,
    "AUC0-24,ss (ng*h/mL)" = auclast,
    "Cav,ss (ng/mL)" = cav
  ) |>
  knitr::kable(digits = 2,
               caption = "Cohort-median steady-state NCA for all three analytes, by enzalutamide arm (PKNCA).")
```

Two qualitative cross-checks against published pharmacology, neither of which
Hadigol 2026 tabulates and so neither of which is a gate:

```{r pharmacology-crosscheck}
enz_cav <- nca_all$cav[nca_all$Analyte == "Enzalutamide" &
                         nca_all$treatment == "160 mg enzalutamide"]
ndm_cav <- nca_all$cav[nca_all$Analyte == "N-desmethyl enzalutamide" &
                         nca_all$treatment == "160 mg enzalutamide"]

# Enzalutamide and its N-desmethyl metabolite circulate at broadly comparable
# steady-state concentrations at 160 mg QD, which is why the paper treats their
# summed concentration as the perpetrator driver. Magnitude bound only.
stopifnot(ndm_cav / enz_cav > 0.4, ndm_cav / enz_cav < 2.5)

tibble::tribble(
  ~Quantity, ~Model, ~`Published expectation`,
  "Enzalutamide Cav,ss at 160 mg QD (ng/mL)", sprintf("%.0f", enz_cav),
  "about 10,000-16,000 (label range; not tabulated in Hadigol 2026)",
  "N-desmethyl / enzalutamide Cav,ss ratio", sprintf("%.2f", ndm_cav / enz_cav),
  "near 1; the two are comparable and equipotent (Methods)"
) |>
  knitr::kable(caption = "Qualitative cross-checks against published enzalutamide pharmacology. Not gated.")
```

# Assumptions and deviations

- **Covariate distributions are assumed.** Hadigol 2026 Tables S1 and S2 (study
  population summary and baseline demographics) are not open access. The virtual
  cohort therefore draws body weight, age and baseline creatinine clearance from
  normal distributions centred on the medians the main text names (80.1 kg, 71
  years) and on the 86.85 mL/min reference, truncated to physiologically
  plausible ranges. No published covariate range, race/ethnicity split or
  regional split could be recorded in the model's `population` metadata. The
  cohort figures are illustrative; they are **not** a reproduction of the
  paper's Figure 3 pvcVPC, whose underlying observations are not available.

- **86.85 mL/min is not claimed to be a median.** The paper introduces it only
  as the baseline creatinine clearance of the typical patient in equation (6).
  With the demographics table unavailable there is no way to confirm it is the
  cohort median, so neither the model file nor this vignette describes it as
  one. By contrast 80.1 kg and 71 years **are** explicitly stated as population
  medians (Table 5 Note).

- **The creatinine-clearance assay is not named.** Hadigol 2026 never states
  which equation produced BCCL. The raw-mL/min renal-disease cutoffs of its
  Table 4 (normal > 90, mild 60-89, moderate 30-59, severe 15-29) are the only
  evidence for the normalisation, and they imply a raw, non-BSA-normalised
  clearance. The model's `covariateData[[CRCL]]` records this explicitly,
  because supplying a BSA-normalised value instead would silently rescale the
  renal effect.

- **Residual error is encoded as log-normal, matching "additive on
  log-transformed data".** All three tables report a "Thetarized sigma" -- the
  NONMEM idiom in which `$SIGMA` is fixed to 1 and the residual SD is carried
  as an estimated THETA -- and the Results state the talazoparib value is "the
  additive residual variability for log-transformed observation data". nlmixr2
  expresses that exactly as `lnorm(expSd)`, so the printed SDs are used
  directly with no CV conversion.

- **No molar-mass correction on metabolite formation.** Hadigol 2026 gives
  `Fmet = 0.634` with no statement about whether it is a mass or molar fraction,
  and the metabolite differs from the parent by one methyl group (about 3% of
  the molecular mass). The model routes `fm` times the enzalutamide mass flux
  into the metabolite compartment, i.e. treats it as a mass fraction, which is
  what reproduces the paper's own numbers. Since `Fmet` was fixed rather than
  estimated and is confounded with `Vcn`, a molar reinterpretation would be
  absorbed into `Vcn` and would not change predicted concentrations.

- **The interaction term is linear and therefore unbounded below.** Equation (1)
  has no Imax; the factor `(1 - 6.58e-6 * (Ce + Cn))` reaches zero at a combined
  perpetrator concentration of about 152,000 ng/mL, roughly five times the
  typical steady-state value at 160 mg QD. That is far outside the observed
  range and the simulated cohort stays well clear of it (minimum factor about
  0.65), but the model asserts `cl > 0` across the cohort rather than assuming
  it, and the form should not be extrapolated to enzalutamide exposures far
  above the therapeutic range.

- **The body-weight exponent on enzalutamide Vce/Fe is 3.495.** This is far
  above any allometric value and makes the central volume very sensitive to
  weight (about 9 L at 60 kg versus about 55 L at 100 kg). It is transcribed as
  printed: Table 1 reports it with SE 0.279, RSE 7.99% and a SIR 95% CI of
  [3.105, 3.904], all of which confirm the magnitude, and the paper notes these
  covariates are not clinically significant and warrant no dose adjustment.

- **Covariates screened but not retained are recorded, not encoded.** Moderate
  CYP3A4 inhibitors (on both CLe/Fe and CLn), moderate P-gp inhibitors (on Ft
  and kat), race (Asian versus non-Asian) and region (Chinese versus
  non-Chinese) were all tested and rejected, and no point estimate is published
  for any of them. They appear in the model's `covariatesDataExcluded` metadata
  so the provenance of the covariate screen survives, without carrying
  convention warnings for declared-but-unused covariates.

- **Absorption is sparsely informed.** Dense sampling was designed only for
  Part 1 (19 of 811 patients); Part 2 used sparse sampling. Food status, a
  covariate on talazoparib kat in the monotherapy model, was not available for
  the combination and the paper flags this as a limitation. Only the hard-capsule
  formulation was used, so the monotherapy model's formulation effect on kat
  does not appear here either.

- **Base models and the sequential-fit intermediates are not extracted.** Tables
  S4 to S8 (base models for each analyte and the model-selection summaries) are
  not open access, and in any case only the final model of each sequential stage
  is a published result. The packaged model carries the three final models as
  one coupled system, which is how the paper uses them.

- **No cross-analyte random-effect covariance.** The three sub-models were fit
  sequentially with upstream parameters fixed to empirical Bayes estimates, so
  no such covariance was estimated and none is invented. The model has three
  independent variability blocks: a 2x2 block for enzalutamide, a 2x2 block for
  the metabolite, and two diagonal etas for talazoparib.

- **All parameter values come from the paper's main-text tables and numbered
  equations.** No value was digitised from a figure, obtained by
  correspondence, or carried from an upstream model file.
