---
title: "Dobutamine in neonates: PK and haemodynamic PD (Hallik 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dobutamine in neonates: PK and haemodynamic PD (Hallik 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Hallik et al. (2020) studied dobutamine in 28 critically ill preterm and term
neonates in their first 72 hours of life. They first fitted a population PK
model to the plasma concentrations alone (Table 2), then fitted six separate
simultaneous PKPD models, one per haemodynamic endpoint (Tables 4 and 5). Each
PKPD fit re-estimated the PK parameters jointly with its endpoint while holding
the maturation parameters fixed at their PK-only values. Following the authors'
structure, the library carries seven models:

| Model | Endpoint | PD form | Source |
|---|---|---|---|
| `Hallik_2020_dobutamine` | plasma concentration only | none | Table 2 |
| `Hallik_2020_dobutamine_rvo` | right ventricular output (`rvo`, mL/kg/min) | linear in plasma C | Table 4 |
| `Hallik_2020_dobutamine_lvo` | left ventricular output (`lvo`, mL/kg/min) | sigmoidal Emax in plasma C | Table 4 |
| `Hallik_2020_dobutamine_lvef` | left ventricular ejection fraction (`lvef`, %) | linear in plasma C | Table 4 |
| `Hallik_2020_dobutamine_hr` | heart rate (`hr`, beats/min) | sigmoidal Emax in effect-site C | Table 5 |
| `Hallik_2020_dobutamine_map` | mean arterial pressure (`map`, mmHg) | sigmoidal Emax in plasma C | Table 5 |
| `Hallik_2020_dobutamine_cftoe` | cerebral fractional tissue O2 extraction (`cftoe`) | sigmoidal Emax in plasma C | Table 5 |

* Citation: `r rxode2::rxode(readModelDb("Hallik_2020_dobutamine"))$reference`
* Description (PK model): `r rxode2::rxode(readModelDb("Hallik_2020_dobutamine"))$description`
* Article: <https://doi.org/10.1111/bcp.14146> (open access, PMC7015735)

# Population

The analysis included 28 of 31 recruited neonates from two Estonian NICUs
(Tallinn Children's Hospital and Tartu University Hospital), April 2016 to
December 2017. Table 1 gives gestational age at birth as median 30.4 weeks
(range 22.7-41.0), with 25% below 28 weeks, 32% at 28-32 weeks, 25% at 32-37
weeks and 18% above 37 weeks; birth weight median 1618 g (465-4380 g); 64%
male; and age at recruitment median 6 h (2-28 h). The main diagnoses were
respiratory distress syndrome, early-onset sepsis, perinatal asphyxia,
meconium aspiration and foeto-foetal transfusion. Dobutamine was infused from
5 ug/kg/min and raised by 5 ug/kg/min about every 30 minutes, up to 20
ug/kg/min. The highest dose reached was 10, 15 and 20 ug/kg/min in 1, 17 and 10
neonates. There were 119 plasma samples, 9 of them below the 0.97 ug/L LLOQ.

The same information is available programmatically via the `population`
element of each model, e.g.
`rxode2::rxode(readModelDb("Hallik_2020_dobutamine"))$population`.

# Source trace

Every `ini()` value carries an in-file comment naming its table row. The
structure shared by all seven models is:

| Equation | Form in the model | Source location |
|---|---|---|
| CL | `exp(lcl + etalcl) * (WT_BIRTH/1.618)^0.75 * PAGE^hill_mat / (tmat50^hill_mat + PAGE^hill_mat)` | Equation 1 |
| V | `exp(lvc + vc_eta_scale * etalcl) * (WT_BIRTH/1.618)` | Equation 2; Methods 2.3 (shared BSV with scale factor on V) |
| Effect site (HR only) | `d/dt(effect) <- ke0 * (Cc - effect)` | Equation 3 |
| Linear PD (RVO, LVEF) | `E = rbase + slope * C` | Equation 4 |
| Sigmoidal Emax PD (LVO, HR, MAP, cFTOE) | `E = rbase + (rmax - rbase) * C^hill / (ec50^hill + C^hill)` | Equation 6 |
| BSV scale | omega2 = (CV/100)^2 | Tables 2, 4, 5 footnote a: CV = sqrt(omega2) x 100% |
| Residual error | proportional on every output | Results 3.1 |

The PK part of each model:

| Model | `lcl` (L/h per 1618 g) | `lvc` (L per 1618 g) | `vc_eta_scale` | `tmat50` (weeks) | `hill_mat` | CL BSV | PK `propSd` | Source |
|---|---|---|---|---|---|---|---|---|
| PK only | 41.2 | 5.29 | 1.34 | 37.4 | 2.67 | 29% | 0.581 | Table 2 |
| RVO | 41.0 | 5.31 | 1.50 | 37.4 fixed | 2.67 fixed | 27% | 0.583 | Table 4 |
| LVO | 40.7 | 5.14 | 1.33 | 37.4 fixed | 2.67 fixed | 25% | 0.589 | Table 4 |
| LVEF | 41.2 | 5.26 | 1.38 | 37.4 fixed | 2.67 fixed | 28% | 0.580 | Table 4 |
| HR | 42.3 | 5.42 | 1.72 | 37.4 fixed | 2.67 fixed | 27% | 0.590 | Table 5 |
| MAP | 37.2 | 4.88 | 3.62 | 37.4 fixed | 2.67 fixed | 24% | 0.675 | Table 5 |
| cFTOE | 37.2 | 4.80 | 2.42 | 37.4 fixed | 2.67 fixed | 35% | 0.653 | Table 5 |

The allometric exponent 0.75 on CL is fixed in every model (Equation 1 prints
it as a literal). The PD part of each model (BSV as CV in parentheses; "-" = not
estimated):

| Model | E0 (`rbase`) | Slope / Emax (`slope` / `rmax`) | EC50 (ug/L) | Hill | keo (1/h) | PD `propSd` | Source |
|---|---|---|---|---|---|---|---|
| RVO | 151 mL/kg/min (41%) | SL 0.214 | - | - | - | 0.184 | Table 4 |
| LVO | 131 mL/kg/min (36%) | Emax 157 mL/kg/min (44%) | 117 | 2.82 | - | 0.167 | Table 4 |
| LVEF | 63.5% (9%) | SL 0.0285 | - | - | - | 0.098 | Table 4 |
| HR | 138 /min (15%) | Emax 172 /min (5%) | 39.2 (50%) | 3.36 | 6.59 | 0.051 | Table 5 |
| MAP | 39.7 mmHg (22%) | Emax 41.9 mmHg (26%) | 25.4 | 13.5 | - | 0.065 | Table 5 |
| cFTOE | 0.227 (50%) | Emax 0.206 (60%) | 52.9 | 3.65 | - | 0.181 | Table 5 |

# Virtual cohort

Individual data are not published. The cohort below reproduces the Table 1
gestational-age bands. Birth weight is drawn around an approximate
50th-percentile birth weight for gestational age (log-normal, SD 0.15 on the
log scale). Postmenstrual age is the gestational age plus 6 hours, the median
age at recruitment.

```{r cohort}
set.seed(2020)
rxode2::rxSetSeed(2020)
n_sub <- 200

# Approximate 50th-percentile birth weight (kg) by gestational age (weeks);
# used only to give the virtual cohort a realistic GA-weight correlation.
bw50 <- data.frame(
  ga = c(22, 24, 26, 28, 30, 32, 34, 36, 38, 40, 42),
  bw = c(0.50, 0.65, 0.90, 1.15, 1.45, 1.80, 2.25, 2.70, 3.10, 3.45, 3.65)
)

make_cohort <- function(n) {
  band <- sample(1:4, n, replace = TRUE, prob = c(0.25, 0.32, 0.25, 0.18))
  lo <- c(22.7, 28, 32, 37)[band]
  hi <- c(28, 32, 37, 41.0)[band]
  ga <- stats::runif(n, lo, hi)
  bw <- stats::approx(bw50$ga, bw50$bw, xout = ga)$y * exp(stats::rnorm(n, 0, 0.15))
  data.frame(
    id = seq_len(n),
    GA = ga,
    WT_BIRTH = pmin(pmax(bw, 0.465), 4.38),
    PAGE = ga + 6 / (24 * 7)
  )
}
cohort <- make_cohort(n_sub)

cohort |>
  summarise(
    `GA median (weeks)` = median(GA),
    `GA range` = paste(round(range(GA), 1), collapse = "-"),
    `Birth weight median (kg)` = median(WT_BIRTH),
    `Birth weight range` = paste(round(range(WT_BIRTH), 2), collapse = "-")
  ) |>
  knitr::kable(digits = 2, caption = "Virtual cohort (compare Table 1: GA 30.4, 22.7-41.0 weeks; BW 1.618, 0.465-4.38 kg).")
```

# Dosing

The simulated regimen follows the study protocol (Methods 2.1): 5 ug/kg/min
from time zero, raised by 5 ug/kg/min every 30 minutes to 20 ug/kg/min, which
is held until 3 h, followed by 1 h of washout. With dose in ug and time in
hours, an infusion of R ug/kg/min enters `central` at a rate of
`R * WT_BIRTH * 60` ug/h.

```{r events}
steps <- data.frame(
  start = c(0, 0.5, 1, 1.5),
  end = c(0.5, 1, 1.5, 3),
  dose_rate = c(5, 10, 15, 20)
)

make_events <- function(cohort, steps, obs_times) {
  doses <- tidyr::crossing(cohort, steps) |>
    mutate(
      time = start,
      evid = 1L,
      cmt = "central",
      rate = dose_rate * WT_BIRTH * 60,
      amt = rate * (end - start)
    ) |>
    select(id, time, evid, cmt, amt, rate, WT_BIRTH, PAGE)
  obs <- tidyr::crossing(cohort, time = obs_times) |>
    mutate(evid = 0L, cmt = "central", amt = NA_real_, rate = NA_real_) |>
    select(id, time, evid, cmt, amt, rate, WT_BIRTH, PAGE)
  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}

obs_times <- sort(unique(c(seq(0, 4, by = 2 / 60), c(0.5, 1, 1.5, 3) - 1 / 60)))
ev <- make_events(cohort, steps, obs_times)
```

# Dobutamine concentrations

## Concentration against infusion rate (Figure 1)

Figure 1 plots each measured concentration against the infusion rate running
at the time. Samples were drawn 15-30 min after each dose change. The
simulation below reads the concentration one minute before each step ends.
The paper's maximum measured concentration was 330 ug/L.

```{r pk-sim}
mod_pk <- rxode2::rxode(readModelDb("Hallik_2020_dobutamine"))
sim_pk <- rxode2::rxSolve(mod_pk, events = ev, keep = c("WT_BIRTH", "PAGE")) |>
  as.data.frame()

step_ends <- data.frame(
  time = c(0.5, 1, 1.5, 3) - 1 / 60,
  dose_rate = c(5, 10, 15, 20)
)
conc_by_rate <- sim_pk |>
  inner_join(step_ends, by = "time")

ggplot(conc_by_rate, aes(factor(dose_rate), Cc)) +
  geom_boxplot(outlier.size = 0.6) +
  geom_hline(yintercept = 330, linetype = "dashed", colour = "grey40") +
  scale_y_log10() +
  labs(
    x = "Infusion rate (ug/kg/min)", y = "Dobutamine (ug/L)",
    title = "Replicates Figure 1 of Hallik 2020",
    caption = "Simulated concentration at the end of each dose step; dashed line = maximum observed (330 ug/L)."
  )

conc_by_rate |>
  group_by(`Infusion rate (ug/kg/min)` = dose_rate) |>
  summarise(
    `Median (ug/L)` = median(Cc),
    `5th pct` = quantile(Cc, 0.05),
    `95th pct` = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 1)
```

The median concentration rises about in proportion to the infusion rate, as
Figure 1 shows for the observed data. The simulated 95th percentile at 20
ug/kg/min is of the same order as the 330 ug/L maximum observed.

## Concentration-time profile (Figure 3)

```{r pk-vpc}
# Drop t = 0 (Cc = 0) only because the axis is logarithmic.
sim_pk[sim_pk$time > 0, ] |>
  group_by(time) |>
  summarise(
    p025 = quantile(Cc, 0.025), p50 = median(Cc), p975 = quantile(Cc, 0.975),
    .groups = "drop"
  ) |>
  ggplot(aes(time, p50)) +
  geom_ribbon(aes(ymin = p025, ymax = p975), alpha = 0.25) +
  geom_line() +
  scale_y_log10() +
  labs(
    x = "Time since start of infusion (h)", y = "Dobutamine (ug/L)",
    title = "Simulated 2.5th, 50th and 97.5th percentiles (compare Figure 3)",
    caption = "Stepped titration 5-10-15-20 ug/kg/min, infusion stopped at 3 h. Figure 3 is prediction-corrected, so its y-axis differs."
  )
```

## Typical-value checks against the text

The Discussion reports that the typical volume of 5.29 L per 1618 g is 3.27
L/kg. The maturation function gives the typical clearance at the cohort's
median postmenstrual age. The mean effect-equilibration time for heart rate
(Results 3.1) is `60 / keo`.

```{r typical-checks}
th <- rxode2::rxode(readModelDb("Hallik_2020_dobutamine"))$theta
fmat <- function(pma, tm50, hill) pma^hill / (tm50^hill + pma^hill)
v_per_kg <- exp(th[["lvc"]]) / 1.618
cl_typ <- exp(th[["lcl"]]) * fmat(30.4, exp(th[["ltmat50"]]), exp(th[["lhill_mat"]]))
thalf_min <- log(2) * exp(th[["lvc"]]) / cl_typ * 60
th_hr <- rxode2::rxode(readModelDb("Hallik_2020_dobutamine_hr"))$theta
mean_eq_time <- 60 / exp(th_hr[["lke0"]])

data.frame(
  Quantity = c(
    "V per kg (L/kg)", "Typical CL at PMA 30.4 weeks (L/h per 1618 g)",
    "Typical CL at PMA 30.4 weeks (mL/min/kg)", "Typical half-life (min)",
    "HR mean equilibration time (min)"
  ),
  Model = c(v_per_kg, cl_typ, cl_typ / 1.618 * 1000 / 60, thalf_min, mean_eq_time),
  Paper = c("3.27 (Discussion 4.1)", "-", "-", "-", "9 (Results 3.1)")
) |>
  knitr::kable(digits = 2)

stopifnot(
  abs(v_per_kg - 3.27) < 0.01,
  abs(mean_eq_time - 9) < 0.5
)
```

# PKNCA validation

To check the PK model with an independent method, three arms of 100 virtual
neonates receive a constant 4-hour infusion of 5, 10 or 20 ug/kg/min, followed
by 2 hours of washout. PKNCA estimates clearance as dose / AUC0-inf. Because
the model is linear, this should recover each subject's own model clearance.
Any difference is trapezoidal integration error, not a model property.

```{r pknca}
nca_cohort <- make_cohort(300) |>
  mutate(treatment = rep(c("5 ug/kg/min", "10 ug/kg/min", "20 ug/kg/min"), each = 100),
         dose_rate = rep(c(5, 10, 20), each = 100))

nca_doses <- nca_cohort |>
  mutate(time = 0, evid = 1L, cmt = "central", rate = dose_rate * WT_BIRTH * 60,
         amt = rate * 4, dur = 4)
nca_obs <- tidyr::crossing(nca_cohort, time = sort(unique(c(seq(0, 6, by = 0.05), 4)))) |>
  mutate(evid = 0L, cmt = "central", amt = NA_real_, rate = NA_real_)
ev_nca <- bind_rows(
  select(nca_doses, id, time, evid, cmt, amt, rate, WT_BIRTH, PAGE, treatment),
  select(nca_obs, id, time, evid, cmt, amt, rate, WT_BIRTH, PAGE, treatment)
) |>
  arrange(id, time, desc(evid))

sim_nca <- rxode2::rxSolve(mod_pk, events = ev_nca, keep = c("treatment"),
                           rtol = 1e-10, atol = 1e-12) |>
  as.data.frame()

conc_nca <- sim_nca |>
  filter(!is.na(Cc)) |>
  mutate(Cc = pmax(Cc, 0)) |>
  select(id, time, Cc, treatment)
conc_nca <- bind_rows(
  conc_nca,
  conc_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(conc_nca, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(
  select(nca_doses, id, time, amt, dur, treatment),
  amt ~ time | treatment + id,
  route = "intravascular", duration = "dur"
)
intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, aucinf.obs = TRUE, cl.obs = TRUE, half.life = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  filter(PPTESTCD %in% c("cmax", "cl.obs", "half.life")) |>
  select(id, treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

model_cl <- sim_nca |>
  group_by(id) |>
  summarise(cl_model = first(cl), vc_model = first(vc), .groups = "drop")

nca_check <- nca_wide |>
  left_join(model_cl, by = "id") |>
  mutate(
    pct_diff_cl = 100 * (cl.obs / cl_model - 1),
    thalf_model = log(2) * vc_model / cl_model
  )

nca_check |>
  group_by(treatment) |>
  summarise(
    `Median Cmax (ug/L)` = median(cmax),
    `Median CL, PKNCA (L/h)` = median(cl.obs),
    `Median CL, model (L/h)` = median(cl_model),
    `Median t1/2, PKNCA (min)` = median(half.life) * 60,
    `Median t1/2, model (min)` = median(thalf_model) * 60,
    .groups = "drop"
  ) |>
  knitr::kable(digits = 2, caption = "PKNCA against each subject's own model parameters, by infusion arm.")

# Same-parameter check: dose/AUC against the subject's own CL, so the only gap
# is trapezoidal error on a 3-minute grid (washout t1/2 ~ 5-40 min).
stopifnot(
  abs(median(nca_check$pct_diff_cl)) < 2,
  quantile(abs(nca_check$pct_diff_cl), 0.9) < 5
)
```

The paper reports no NCA of its own data, so there is no published table to
compare against. Clearance does not change with the infusion rate, which
matches the paper's finding that dobutamine PK was linear over 5-20 ug/kg/min
(Results 3.1).

# Haemodynamic endpoints

## Typical concentration-effect curves

Each endpoint's typical-value concentration-effect relationship, evaluated
over the observed concentration range (up to 330 ug/L). For heart rate the
driver is the effect-site concentration, which equals the plasma concentration
at steady state.

```{r ce-curves}
theta_of <- function(name) rxode2::rxode(readModelDb(name))$theta
emax_curve <- function(conc, th, sfx) {
  e0 <- exp(th[[paste0("lrbase_", sfx)]])
  emax <- exp(th[[paste0("lrmax_", sfx)]])
  ec50 <- exp(th[[paste0("lec50_", sfx)]])
  hill <- exp(th[[paste0("lhill_", sfx)]])
  e0 + (emax - e0) * conc^hill / (ec50^hill + conc^hill)
}
linear_curve <- function(conc, th, sfx) {
  exp(th[[paste0("lrbase_", sfx)]]) + exp(th[[paste0("lslope_", sfx)]]) * conc
}

conc_grid <- seq(0, 330, by = 1)
ce <- bind_rows(
  data.frame(endpoint = "RVO (mL/kg/min)", conc = conc_grid,
             effect = linear_curve(conc_grid, theta_of("Hallik_2020_dobutamine_rvo"), "rvo")),
  data.frame(endpoint = "LVO (mL/kg/min)", conc = conc_grid,
             effect = emax_curve(conc_grid, theta_of("Hallik_2020_dobutamine_lvo"), "lvo")),
  data.frame(endpoint = "LVEF (%)", conc = conc_grid,
             effect = linear_curve(conc_grid, theta_of("Hallik_2020_dobutamine_lvef"), "lvef")),
  data.frame(endpoint = "HR (beats/min)", conc = conc_grid,
             effect = emax_curve(conc_grid, theta_of("Hallik_2020_dobutamine_hr"), "hr")),
  data.frame(endpoint = "MAP (mmHg)", conc = conc_grid,
             effect = emax_curve(conc_grid, theta_of("Hallik_2020_dobutamine_map"), "map")),
  data.frame(endpoint = "cFTOE", conc = conc_grid,
             effect = emax_curve(conc_grid, theta_of("Hallik_2020_dobutamine_cftoe"), "cftoe"))
)

ggplot(ce, aes(conc, effect)) +
  geom_line() +
  facet_wrap(~endpoint, scales = "free_y") +
  labs(x = "Dobutamine concentration (ug/L)", y = "Typical endpoint value",
       title = "Typical concentration-effect relationships (Tables 4 and 5)")
```

The Discussion makes several claims about these curves. Mean heart rate rises
from 138 to 172 beats/min. The MAP and HR effects reach their maximum "within
concentrations of 50 and 80 ug/L, respectively". LVO keeps rising "at least up
to concentration of 200 ug/L". cFTOE falls with dobutamine. The fraction of
each maximal change reached at those concentrations is checked below.

```{r ce-checks}
frac_of_max <- function(conc, th, sfx) {
  e0 <- exp(th[[paste0("lrbase_", sfx)]])
  (emax_curve(conc, th, sfx) - e0) / (exp(th[[paste0("lrmax_", sfx)]]) - e0)
}
th_map <- theta_of("Hallik_2020_dobutamine_map")
th_lvo <- theta_of("Hallik_2020_dobutamine_lvo")
th_cftoe <- theta_of("Hallik_2020_dobutamine_cftoe")

checks <- data.frame(
  Claim = c(
    "HR baseline -> maximum (beats/min)",
    "MAP: fraction of maximal change at 50 ug/L",
    "HR: fraction of maximal change at 80 ug/L",
    "LVO: fraction of maximal change at 200 ug/L",
    "cFTOE change at 330 ug/L (typical)"
  ),
  Model = c(
    sprintf("%.0f -> %.0f", exp(th_hr[["lrbase_hr"]]), exp(th_hr[["lrmax_hr"]])),
    sprintf("%.3f", frac_of_max(50, th_map, "map")),
    sprintf("%.3f", frac_of_max(80, th_hr, "hr")),
    sprintf("%.3f", frac_of_max(200, th_lvo, "lvo")),
    sprintf("%+.3f", emax_curve(330, th_cftoe, "cftoe") - exp(th_cftoe[["lrbase_cftoe"]]))
  )
)
knitr::kable(checks)

stopifnot(
  frac_of_max(50, th_map, "map") > 0.99,
  frac_of_max(80, th_hr, "hr") > 0.9,
  frac_of_max(200, th_lvo, "lvo") < 0.9,
  emax_curve(330, th_cftoe, "cftoe") < exp(th_cftoe[["lrbase_cftoe"]])
)
```

## Endpoint time courses under the titration (Figure 4)

Figure 4 shows prediction-corrected VPCs of each endpoint over time. The
simulation below applies the stepped titration to the virtual cohort for each
of the six PKPD models. It shows the median and 95% interval of each endpoint.
Each PKPD model has two outputs, the concentration and the endpoint, and
neither is an ODE state. The observation rows are therefore keyed by
`dvid = 1` with no compartment, and rxode2 returns both output columns at every
observation time.

```{r pd-sim}
pd_models <- c(
  rvo = "Hallik_2020_dobutamine_rvo", lvo = "Hallik_2020_dobutamine_lvo",
  lvef = "Hallik_2020_dobutamine_lvef", hr = "Hallik_2020_dobutamine_hr",
  map = "Hallik_2020_dobutamine_map", cftoe = "Hallik_2020_dobutamine_cftoe"
)
pd_labels <- c(
  rvo = "RVO (mL/kg/min)", lvo = "LVO (mL/kg/min)", lvef = "LVEF (%)",
  hr = "HR (beats/min)", map = "MAP (mmHg)", cftoe = "cFTOE"
)

ev_pd <- ev |>
  mutate(
    dvid = ifelse(evid == 0L, 1L, NA_integer_),
    cmt = ifelse(evid == 0L, NA_character_, cmt)
  )

sim_pd <- lapply(names(pd_models), function(out) {
  mod <- rxode2::rxode(readModelDb(pd_models[[out]]))
  s <- rxode2::rxSolve(mod, events = ev_pd, useLinCmt = FALSE) |> as.data.frame()
  data.frame(id = s$id, time = s$time, endpoint = pd_labels[[out]], value = s[[out]])
}) |>
  bind_rows()

pd_summary <- sim_pd |>
  group_by(endpoint, time) |>
  summarise(
    p025 = quantile(value, 0.025), p50 = median(value), p975 = quantile(value, 0.975),
    .groups = "drop"
  )

ggplot(pd_summary, aes(time, p50)) +
  geom_ribbon(aes(ymin = p025, ymax = p975), alpha = 0.25) +
  geom_line() +
  geom_vline(xintercept = c(0.5, 1, 1.5, 3), linetype = "dotted", colour = "grey50") +
  facet_wrap(~endpoint, scales = "free_y") +
  labs(x = "Time since start of infusion (h)", y = "Endpoint value",
       title = "Simulated endpoint time courses (compare Figure 4)",
       caption = "Dotted lines: dose steps at 0.5, 1, 1.5 h and infusion stop at 3 h.")
```

```{r pd-checks}
# Median endpoint at baseline and at the end of the 20 ug/kg/min step.
pd_table <- sim_pd |>
  filter(time %in% c(0, 3 - 1 / 60)) |>
  mutate(when = ifelse(time == 0, "Baseline", "End of 20 ug/kg/min")) |>
  group_by(endpoint, when) |>
  summarise(median = median(value), .groups = "drop") |>
  tidyr::pivot_wider(names_from = when, values_from = median)
knitr::kable(pd_table, digits = 3,
             caption = "Median simulated endpoint before the infusion and at the top dose.")

hr_top <- pd_table$`End of 20 ug/kg/min`[pd_table$endpoint == "HR (beats/min)"]
hr_base <- pd_table$Baseline[pd_table$endpoint == "HR (beats/min)"]
# Baseline medians are the E0 estimates (lognormal BSV: median = typical).
# Robust centre checks only (see the article's assumptions section).
stopifnot(
  abs(hr_base / 138 - 1) < 0.05,
  hr_top > hr_base,
  hr_top <= 172 * 1.05
)
```

# Assumptions and deviations

* **Model count.** The paper reports one PK-only model and six separately fitted
  PKPD models, each with its own re-estimated PK parameters. All seven are
  carried as separate models, so a user picks the fit that matches the
  endpoint of interest. The six PKPD models describe the same drug but are not
  a single joint model.
* **Emax is a plateau level.** Equations 5 and 6 write
  `E = E0 + (Emax - E0) * ...`. The paper defines Emax as "the estimated
  maximum HD parameter value", so it is the plateau level, not the increment.
  The models therefore name it `rmax_<endpoint>` rather than `emax`, following
  `Dings_2026_cafedrine_theodrenaline_ephedrine`. For cFTOE the plateau (0.206)
  lies below the baseline (0.227), so the effect is a fall, as the Discussion
  states.
* **Shared random effect.** Methods: "a shared BSV was used with an estimated
  scale factor applied for V". The single eta is placed on CL and enters V
  multiplied by `vc_eta_scale`. The tables print the same BSV value in the CL
  and V rows. We read that as the variance of the shared eta, so the implied
  CV of V is the scale factor times the CL CV (for example 1.34 x 29% = 39% in
  the PK-only model).
* **V exponent.** Equation 2 prints no exponent on (BW/Wst), so V scales
  linearly with birth weight.
* **Birth weight as the size covariate.** The paper scales on birth weight. The
  models use `WT_BIRTH` in kg with a 1.618 kg reference, which is the same
  ratio as the paper's 1618 g. Infusion rates in ug/kg/min are converted with
  the same weight.
* **PMA in weeks.** `PAGE` is declared in weeks, as the paper states it,
  rather than the register's default of months.
* **Residual error.** Every residual error is read as a proportional SD, not a
  variance. This fits the Results statement that residual variability was
  ">50% in PK observations and between 5-20% in PD observations".
* **BLQ handling.** Concentrations below the 0.97 ug/L LLOQ were set to 0.5
  ug/L in the fit (Methods 2.1). This affects estimation only, not the
  simulation model.
* **Covariates screened and not retained** (Methods 2.3): postnatal age,
  antenatal glucocorticoids, dopamine co-administration, haemoglobin, albumin,
  patent ductus arteriosus diameter, baseline LVEF and baseline RVO. Those with
  a canonical column are listed in `covariatesDataExcluded` of the PK-only
  model.
* **Virtual cohort.** Individual data are not published. The gestational-age
  bands follow Table 1. Birth weight for gestational age uses an approximate
  50th-percentile curve with 15% log-scale spread. This is the maintainers'
  approximation and only affects the spread of the simulated profiles, not the
  typical-value checks.
* **No published NCA.** The paper gives no Cmax, AUC or half-life summaries.
  The PKNCA section therefore checks the model against itself (dose/AUC
  against each subject's clearance) rather than against published values.
* **Errata.** A search of Europe PMC and Crossref (September 2026) found no
  erratum or correction for this article.
