---
title: "Acetaminophen (Hammer 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Acetaminophen (Hammer 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Hammer_2020_acetaminophen"))$reference`
* Description: `r rxode2::rxode(readModelDb("Hammer_2020_acetaminophen"))$description`
* Article: <https://doi.org/10.1002/jcph.1508> (open access, PMC6973014)

# Population

Hammer 2020 ran a randomized, placebo-controlled multicentre US study
(NCT01635101) of intravenous acetaminophen in surgical neonates and infants
under 24 months of age with acute postoperative pain. Subjects were stratified
as neonates (< 28 days; extreme preterm, preterm and full term by gestational
age), younger infants (28 days to < 6 months), intermediate-age infants (6 to
< 12 months) and older infants (12 to < 24 months). They were randomized
2:2:1:1 to a low dose (7.5 mg/kg in extreme preterm neonates up to 12.5 mg/kg
in infants), a high dose (10 to 15 mg/kg), or saline placebo (two control
groups), each given as a 15-minute IV infusion every 6 h for 4 doses, with
standard-of-care opioids.

The efficacy population (Table 2, n = 197) had a median postnatal age of
195 days and a median screening weight of 7.0 kg (range 0.8 to 14.3 kg), and
was 35.5% female and 68.5% White. The PK population was 581 samples from 158
subjects. The final model was fitted to these data combined with the 1260
samples from 125 pediatric subjects (Zuppa and Palmer studies, neonates to
adolescents) used for the previously developed model. The pooled data span
about 33.7 to 906 weeks of postmenstrual age (Figure 4).

The same information is available programmatically via
`readModelDb("Hammer_2020_acetaminophen")()$population`.

# Source trace

Every `ini()` value carries an in-file comment pointing to its source in
`inst/modeldb/specificDrugs/Hammer_2020_acetaminophen.R`. All values come from
Table 3 (page 20), 'Current Model' column.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` | log(18.9) L/h | Table 3: CL = 18.9 x (WT/70)^0.75 |
| `lvc` | log(23.0) L | Table 3: Vc = 23.0 x (WT/70) |
| `lq` | log(47.7) L/h | Table 3: CLp = 47.7 x (WT/70)^0.75 |
| `lvp` | log(45.5) L | Table 3: Vp = 45.5 x (WT/70) |
| `e_wt_cl_q` | 0.75 (fixed) | Table 3 exponents on CL and CLp |
| `e_wt_vc_vp` | 1 (fixed) | Table 3 linear weight on Vc and Vp |
| `e_placebo_cl` | 0.524 | Table 3: 'Placebo on CL' x0.524 |
| `e_page_cl` | 0.611 | Table 3 maturation function: 1 - 0.611 x exp(...) |
| `thalf_cl` | 32.6 weeks (fixed) | Table 3 maturation function, ln(2)/32.6 with footnote 'Fixed at previous value' |
| `etalcl`, `etalvc` | 0.127, 0.993; cov 0.18253 | Table 3 omega^2 CL, omega^2 Vc; note 'correlation between CL and Vc BSV was 51.4%' |
| `expSd` | 0.221 | Table 3: 'Log residual error' |
| `fmat = 1 - e_page_cl * exp(-(PAGE - 40) * log(2) / thalf_cl)` | n/a | Table 3 maturation function |
| `cl = ... * e_placebo_cl^PLACEBO` | n/a | Table 3 'Placebo on CL' row |
| `d/dt(central)`, `d/dt(peripheral1)` | n/a | Methods: 2-compartment model with linear elimination |
| `Cc ~ lnorm(expSd)` | n/a | Table 3: 'Log residual error' |

# Virtual cohort

The observed data are not public. The first-dose comparison below uses one
virtual arm per row of Table 4 (subgroup by dose level). Table 4 prints each
subgroup's median CL both in L/h and in L/h/kg, so the median body weight of
each subgroup is their ratio. The virtual subjects draw weight log-normally
around that median (SD 0.15 on the log scale) and postmenstrual age uniformly
over a range that fits the subgroup's gestational and postnatal age definition.

```{r cohort}
set.seed(2020)

# Median weight of each Table 4 subgroup = median CL (L/h) / median CL (L/h/kg).
subgroups <- tibble::tribble(
  ~subgroup, ~cl_lh, ~cl_lhkg, ~page_lo, ~page_hi,
  "Extreme preterm neonates", 0.238, 0.192, 29, 33,
  "Preterm neonates", 0.465, 0.220, 33, 39,
  "Full-term neonates", 0.673, 0.217, 37.5, 44,
  "Younger infants", 1.58, 0.299, 44, 66,
  "Intermediate infants", 2.46, 0.336, 66, 92,
  "Older infants", 3.83, 0.368, 92, 144
) |>
  mutate(wt_med = cl_lh / cl_lhkg)

# Table 4 dose levels (mg/kg) by subgroup. The extreme preterm and preterm
# neonates only report a high-dose arm.
arms <- tibble::tribble(
  ~subgroup, ~level, ~mgkg,
  "Extreme preterm neonates", "High", 10,
  "Preterm neonates", "High", 12.5,
  "Full-term neonates", "Low", 10,
  "Full-term neonates", "High", 12.5,
  "Younger infants", "Low", 12.5,
  "Younger infants", "High", 15,
  "Intermediate infants", "Low", 12.5,
  "Intermediate infants", "High", 15,
  "Older infants", "Low", 12.5,
  "Older infants", "High", 15
) |>
  left_join(subgroups, by = "subgroup") |>
  mutate(treatment = paste0(subgroup, ", ", level, " (", mgkg, " mg/kg)"))

n_per_arm <- 150
obs_times <- c(0, 0.1, 0.2, 0.25, 0.35, 0.5, 0.75, 1, 1.5, 2, 3, 4, 5, 6)

subjects <- arms |>
  slice(rep(seq_len(n()), each = n_per_arm)) |>
  mutate(
    id = seq_len(n()),
    WT = wt_med * exp(rnorm(n(), 0, 0.15)),
    PAGE = runif(n(), page_lo, page_hi),
    PLACEBO = 0,
    amt = mgkg * WT
  )

dose_rows <- subjects |>
  transmute(
    id, treatment, WT, PAGE, PLACEBO,
    time = 0, evid = 1, amt, dur = 0.25, cmt = "central"
  )
obs_rows <- subjects |>
  select(id, treatment, WT, PAGE, PLACEBO) |>
  tidyr::crossing(time = obs_times) |>
  mutate(evid = 0, amt = 0, dur = 0, cmt = "central")
events <- bind_rows(dose_rows, obs_rows) |>
  arrange(id, time, desc(evid))
stopifnot(max(table(subjects$treatment)) <= 200)
```

# Simulation

```{r simulate}
mod <- readModelDb("Hammer_2020_acetaminophen")
sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep = c("treatment", "WT"),
  returnType = "data.frame"
)
```

# Replicate published figures

## Figure 4: clearance versus postmenstrual age

Figure 4 shows the sigmoidal rise of clearance, standardized to 70 kg, towards
a plateau of 18.9 L/h in children and adolescents. The typical curve follows
directly from the Table 3 maturation function.

```{r figure-4}
e <- rxode2::rxode(mod)$theta
fig4 <- tibble(PAGE = seq(30, 906, by = 2)) |>
  mutate(
    fmat = 1 - e[["e_page_cl"]] * exp(-(PAGE - 40) * log(2) / e[["thalf_cl"]]),
    Active = exp(e[["lcl"]]) * fmat,
    Placebo = Active * e[["e_placebo_cl"]]
  ) |>
  tidyr::pivot_longer(c(Active, Placebo), names_to = "Arm", values_to = "cl70")

ggplot(fig4, aes(PAGE, cl70, colour = Arm)) +
  geom_line() +
  geom_hline(yintercept = exp(e[["lcl"]]), linetype = "dashed") +
  labs(
    x = "Postmenstrual age (weeks)",
    y = "Typical CL standardized to 70 kg (L/h)",
    caption = "Replicates the typical-value curve of Figure 4 of Hammer 2020."
  )
```

```{r maturation_milestones}
# PMA at which fmat reaches a fraction f of the adult value:
# 1 - 0.611 * exp(-(PMA - 40) ln2 / 32.6) = f
pma_at <- function(f) 40 + e[["thalf_cl"]] * log(e[["e_page_cl"]] / (1 - f)) / log(2)
milestones <- c(`fmat at 40 weeks` = 1 - e[["e_page_cl"]], `PMA at 50%` = pma_at(0.5), `PMA at 90%` = pma_at(0.9))
round(milestones, 2)
```

The maturation function starts at `r round(100 * milestones[[1]])`% of the
adult value at 40 weeks, passes 50% at about `r round(milestones[[2]])` weeks and
90% at about `r round(milestones[[3]])` weeks (about 1.6 years postnatal). This
matches Figure 4, which shows clearance levelling off from roughly 2 years
onwards.

## First-dose profiles by subgroup

```{r profiles}
# Drop the pre-dose sample only for the log-scale plot (PKNCA below keeps it).
sim[sim$time != 0, ] |>
  group_by(treatment, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = median(Cc),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~treatment, ncol = 3) +
  scale_y_log10() +
  labs(
    x = "Time after first dose (h)",
    y = "Acetaminophen Cc (ug/mL)",
    caption = "Median and 90% prediction interval, 150 virtual subjects per arm."
  )
```

# PKNCA validation

Table 4 reports, per subgroup and dose level, the median Cmax and AUCtau of the
first dose (tau = 6 h), derived from the individual post hoc parameters. The
same quantities are computed here with PKNCA from the virtual cohort.

```{r pknca}
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(
  events |> filter(evid == 1) |> select(id, time, amt, treatment),
  amt ~ time | treatment + id
)
intervals <- data.frame(start = 0, end = 6, cmax = TRUE, auclast = TRUE)
nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)
```

## Comparison against Table 4

```{r nca_compare}
published <- tibble::tribble(
  ~treatment, ~cmax, ~auclast,
  "Preterm neonates, High (12.5 mg/kg)", 22.8, 43.2,
  "Full-term neonates, Low (10 mg/kg)", 18.3, 34.3,
  "Full-term neonates, High (12.5 mg/kg)", 22.6, 44.6,
  "Younger infants, Low (12.5 mg/kg)", 21.6, 34.8,
  "Younger infants, High (15 mg/kg)", 26.7, 41.2,
  "Intermediate infants, Low (12.5 mg/kg)", 20.8, 30.0,
  "Intermediate infants, High (15 mg/kg)", 29.1, 41.6,
  "Older infants, Low (12.5 mg/kg)", 21.2, 27.2,
  "Older infants, High (15 mg/kg)", 27.9, 38.7
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "treatment",
  units = c(cmax = "ug/mL", auclast = "ug*h/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Median first-dose NCA: Hammer 2020 Table 4 vs simulation.")
```

The simulated median first-dose Cmax and AUC0-6 match the Table 4 medians in
every reported arm within the 20% flagging tolerance (typically within about
10%). The subgroup weights and postmenstrual ages are approximations of cohorts
that are not tabulated, so this agreement also covers the maturation function
and the allometric scaling across a 10-fold weight range.

```{r nca_gate}
sim_med <- as.data.frame(nca_res$result) |>
  filter(PPTESTCD %in% c("cmax", "auclast")) |>
  group_by(treatment, PPTESTCD) |>
  summarise(sim = median(PPORRES), .groups = "drop")
chk <- published |>
  tidyr::pivot_longer(c(cmax, auclast), names_to = "PPTESTCD", values_to = "ref") |>
  inner_join(sim_med, by = c("treatment", "PPTESTCD")) |>
  mutate(pct_diff = 100 * (sim - ref) / ref)
stopifnot(nrow(chk) == 18)
summary(chk$pct_diff)
stopifnot(
  # Structural: a mis-transcribed CL, volume or maturation constant shifts
  # every arm by tens of percent.
  abs(median(chk$pct_diff)) < 10,
  # Envelope: robust to which virtual subjects land in the tails.
  quantile(abs(chk$pct_diff), 0.9) < 25
)
```

## Typical CL, Vss and terminal half-life by subgroup

Table 4 also reports the median post hoc CL (L/h/kg), Vss (L/kg) and terminal
half-life. For the median weight of each subgroup and the midpoint of its
postmenstrual-age range, the typical values follow in closed form from the
model: Vss = Vc + Vp, and the terminal half-life is ln 2 over the smaller
eigenvalue of the two-compartment system.

```{r typical_values}
typ <- subgroups |>
  mutate(
    PAGE = (page_lo + page_hi) / 2,
    fmat = 1 - e[["e_page_cl"]] * exp(-(PAGE - 40) * log(2) / e[["thalf_cl"]]),
    cl = exp(e[["lcl"]]) * (wt_med / 70)^0.75 * fmat,
    vc = exp(e[["lvc"]]) * wt_med / 70,
    vp = exp(e[["lvp"]]) * wt_med / 70,
    q = exp(e[["lq"]]) * (wt_med / 70)^0.75,
    k10 = cl / vc,
    k12 = q / vc,
    k21 = q / vp,
    beta = 0.5 * ((k10 + k12 + k21) - sqrt((k10 + k12 + k21)^2 - 4 * k10 * k21)),
    `CL model (L/h/kg)` = cl / wt_med,
    `Vss model (L/kg)` = (vc + vp) / wt_med,
    `t1/2 beta model (h)` = log(2) / beta
  ) |>
  left_join(
    tibble::tribble(
      ~subgroup, ~`Vss Table 4 (L/kg)`, ~`t1/2 beta Table 4 (h)`,
      "Extreme preterm neonates", 0.848, 3.26,
      "Preterm neonates", 0.922, 3.23,
      "Full-term neonates", 0.916, 3.38,
      "Younger infants", 0.948, 2.71,
      "Intermediate infants", 0.926, 2.29,
      "Older infants", 0.952, 2.15
    ),
    by = "subgroup"
  ) |>
  mutate(`CL Table 4 (L/h/kg)` = cl_lhkg)

typ |>
  select(
    Subgroup = subgroup,
    `PMA (weeks)` = PAGE,
    `Weight (kg)` = wt_med,
    `CL Table 4 (L/h/kg)`, `CL model (L/h/kg)`,
    `Vss Table 4 (L/kg)`, `Vss model (L/kg)`,
    `t1/2 beta Table 4 (h)`, `t1/2 beta model (h)`
  ) |>
  knitr::kable(digits = 3)

# Closed-form, no random draws: a structural transcription error moves these
# by far more than the bounds.
cl_diff <- with(typ, `CL model (L/h/kg)` / `CL Table 4 (L/h/kg)` - 1)
hl_diff <- with(typ, `t1/2 beta model (h)` / `t1/2 beta Table 4 (h)` - 1)
stopifnot(
  all(abs(cl_diff) < 0.25),
  all(abs(hl_diff) < 0.30),
  all(abs(typ$`Vss model (L/kg)` / typ$`Vss Table 4 (L/kg)` - 1) < 0.20)
)
```

The typical CL per kilogram rises from about 0.19 L/h/kg in extreme preterm
neonates to about 0.37 L/h/kg in older infants, as in Table 4. The model's
typical Vss is 68.5 L per 70 kg (0.979 L/kg) at every age, 3% to 15% above the
Table 4 medians of 0.85 to 0.95 L/kg. Those medians summarize the post hoc
(empirical Bayes) estimates of the observed subjects rather than the typical
value, so a small offset is expected. The typical terminal
half-life falls from about 3.2 h in neonates to about 2.2 h in older infants,
as in Table 4.

# Assumptions and deviations

* **Residual error scale.** Table 3 prints a single 'Log residual error' of
  0.221 in a table whose IIV rows are labelled as variances (omega^2). It is
  encoded as the log-scale SD (`expSd = 0.221`, about 22% CV), not as a
  variance (SD 0.470). The Figure 2 prediction-corrected VPC separates the two
  readings. At 1 to 2 h after the dose its simulated 95th percentile is only
  about 1.5 times the median, while a residual SD of 0.470 alone would put
  it at least exp(1.645 x 0.470) = 2.2 times the median, before adding any
  between-subject variability. Simulating the model with SD 0.221 gives about
  1.7 to 1.8. The previous model's 'sigma^2 prop (%) 27.9' is also clearly a CV
  rather than a variance, so the table's sigma labels are not reliable.
* **omega^2 Vc = 0.993** is taken as a variance, as labelled. Its square root
  (0.9965) is almost the same number, so the SD reading would give nearly the
  same model.
* **Maturation half-life fixed.** The 32.6-week half-life carries the Table 3
  footnote 'Fixed at previous value derived in.', and the citation is missing
  from the footnote. The value is encoded as `fixed(32.6)`, and 0.611 as
  estimated. The Results prose mentions 're-estimation of the constant
  describing age-related changes in CL', which is taken to refer to 0.611.
  The previous-model column prints 41.0 weeks, so 32.6 is not that column's
  value.
* **Placebo multiplier on CL.** The placebo control subjects received saline
  study drug but had residual acetaminophen concentrations (Figure 1). The
  paper estimates their CL as 0.524 times the active-arm value, but does not
  describe the dosing that produced those concentrations. The effect is kept
  as the `PLACEBO` covariate. For the approved IV regimen, simulate with
  `PLACEBO = 0`.
* **No IIV on CLp or Vp.** The current model reports IIV on CL and Vc only
  (the previous model also had it on CLp and Vp).
* **Previous model not extracted.** The 'Previous Model' column of Table 3 (the
  Zuppa and Palmer pediatric model) is the predecessor, reported for
  comparison. Its residual-error rows ('sigma^2 prop (%) 27.9', 'sigma^2 add
  (mg/L) 168.2') cannot be encoded without guessing units, and the paper cites
  no publication for it.
* **PK-PD not extracted.** The effect-compartment equilibration half-life was
  'fixed to the value estimated' in a cited reference and is not printed.
  The pain-score and opioid-rescue analyses are linear regressions and a
  survival analysis with no reported parameters.
* **Table 4 extreme-preterm high-dose cells.** The Cmax and AUCtau printed for
  extreme preterm neonates (20.19 (6.259), 21.2 (4.37-30.9) ug/mL; 27.70
  (8.092), 27.2 (14.1-46.3) ug.h/mL) are identical, digit for digit, to the
  older-infant low-dose cells. They also report an SD and a range that a
  2-subject group could not produce, so they appear to be a copy error in the
  paper and are left out of the NCA comparison. The extreme preterm subgroup
  is still simulated, and its CL, Vss and half-life are compared.
* **Virtual covariates.** Subgroup median weights are derived from Table 4
  (median CL in L/h divided by median CL in L/h/kg). The postmenstrual-age
  ranges are assumptions consistent with each subgroup's definition, because
  the paper does not tabulate PMA by subgroup.
* **Errata.** A EuropePMC search (2026-09-25) found no correction notice for
  doi:10.1002/jcph.1508.
