---
title: "Fluconazole (Han 2013)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fluconazole (Han 2013)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r meta}
mod_meta <- nlmixr2est::nlmixr(readModelDb("Han_2013_fluconazole"))$meta
```

* Citation: `r mod_meta$reference`
* Description: `r mod_meta$description`
* Article (DOI): <https://doi.org/10.1128/AAC.01372-12>

This vignette validates the packaged `Han_2013_fluconazole` model -- a
one-compartment IV population PK model for fluconazole in 60 adult
burn-ICU patients with suspected or confirmed Candida infection -- against
the source publication's Figure 2 (steady-state VPC by 100, 200, and 400
mg daily regimens) and Figure 3 (probability of fAUC/MIC target attainment
across patient subtypes at 200 and 400 mg/day). The packaged model
encodes the piecewise CL covariate structure of Han 2013 Table 3, in which
a fixed CRRT-cohort CL (`theta3 = 1.85 L/h`) substitutes for the additive
non-CRRT formula (`theta1 + (CLCR/120)*theta4 + TBI*theta7 + SEPS*theta9`)
whenever `CRRT_STATUS = 1`, and an additive WT / edema / postburn-recency
formula on V (`theta5*WT/65 + theta6*EDEM + theta8*TBI`) without the
not-estimable `theta2` intercept.

# Population

Han 2013 enrolled 60 burn-ICU patients (50 male / 10 female) at Hangang
Sacred Heart Hospital between December 2008 and May 2010, with burns
ranging from 11 to 95 percent of total body surface area (mean TBSA
50.3 percent). Patients were intravenously infused with 100-400 mg
fluconazole over 10-40 minutes every 24 h for treatment of suspected
(47/60) or confirmed (13/60) fungal infection, and PK sampling was
performed at steady state after 5-17 days (mean 6.5 days) of fluconazole
therapy. The cohort had mean age 50.3 years (range 20-82), mean weight
65.6 kg (range 40-90 kg), mean Cockcroft-Gault creatinine clearance
123.5 mL/min (range 21.6-282.7), and APACHE II score mean 7.6 (range
2-26). Twenty-nine of 60 had clinically diagnosed sepsis, 20/60 had
clinical edema (puffy face and pitting peripheral edema), 15/60 were on
continuous renal replacement therapy, and mean time from burn injury to
fluconazole administration was 23 days (range 7-88) (Han 2013 Table 1).

The same information is available programmatically via the model's
`population` metadata:

```{r population}
str(mod_meta$population)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Han_2013_fluconazole.R`. The
table below collects them in one place; all values come from Han 2013
Table 3 (Final parameter estimates and bootstrap results).

| Parameter / equation                                                                                  | Value                          | Source location                       |
|-------------------------------------------------------------------------------------------------------|--------------------------------|---------------------------------------|
| `lcl` (CL intercept, non-CRRT)                                                                        | `log(0.693)`                   | Table 3 row `theta1` = 0.693 L/h      |
| `lcl_rrt` (CL on CRRT)                                                                                | `log(1.85)`                    | Table 3 row `theta3` = 1.85 L/h       |
| `lvc` (V proportionality at WT = 65 kg)                                                               | `log(50.3)`                    | Table 3 row `theta5` = 50.3 L         |
| `e_crcl_cl` (CL slope per CLCR/120)                                                                   | 0.557                          | Table 3 row `theta4` = 0.557 L/h      |
| `e_dis_burn_recent_cl` (CL shift if within 30 postburn days)                                          | 0.504                          | Table 3 row `theta7` = 0.504 L/h      |
| `e_dis_sepsis_cl` (CL shift if sepsis)                                                                | -0.369                         | Table 3 row `theta9` = -0.369 L/h     |
| `e_dis_edema_vc` (V shift if edema)                                                                   | 13.6                           | Table 3 row `theta6` = 13.6 L         |
| `e_dis_burn_recent_vc` (V shift if within 30 postburn days)                                           | 9.61                           | Table 3 row `theta8` = 9.61 L         |
| `etalcl ~ 0.12000`                                                                                    | `log(0.357^2 + 1)`             | Table 3 row `omega1^2` BSV CL non-CRRT = 35.7 %CV |
| `etalvc ~ 0.02983`                                                                                    | `log(0.174^2 + 1)`             | Table 3 row `omega2^2` BSV V = 17.4 %CV |
| `etalcl_rrt ~ 0.02559`                                                                                | `log(0.161^2 + 1)`             | Table 3 row `omega3^2` BSV CL CRRT = 16.1 %CV |
| `propSd <- 0.3332`                                                                                    | `sqrt(0.111)`                  | Table 3 row `sigma1^2` = 0.111 (proportional variance) |
| `cl_norrt <- (exp(lcl) + e_crcl_cl*(CRCL/120) + e_dis_burn_recent_cl*DIS_BURN_RECENT + e_dis_sepsis_cl*DIS_SEPSIS) * exp(etalcl)` | n/a | Table 3 caption: `CL = (1 - CRRT)*(theta1 + CLCR*theta4 + TBI*theta7 + SEPS*theta9) + CRRT*theta3`; Table 3 `theta4` description "Proportionality constant between CL and CLCR/120" fixes the (CLCR/120) normalisation. |
| `cl_rrt <- exp(lcl_rrt) * exp(etalcl_rrt)`                                                            | n/a                            | Table 3 caption: `+ CRRT*theta3`      |
| `cl <- (1 - CRRT_STATUS)*cl_norrt + CRRT_STATUS*cl_rrt`                                               | n/a                            | Table 3 caption                       |
| `vc <- ((WT/65)*exp(lvc) + e_dis_edema_vc*DIS_EDEMA + e_dis_burn_recent_vc*DIS_BURN_RECENT) * exp(etalvc)` | n/a                       | Table 3 caption: `V = theta2 + (WT/65)*theta5 + EDEM*theta6 + TBI*theta8`; Table 3 footnote b: theta2 "not successfully estimated" -> dropped |
| `d/dt(central) <- -kel*central`                                                                       | n/a                            | Han 2013 Results, "one-compartment first-order elimination model" |
| `Cc ~ prop(propSd)`                                                                                   | n/a                            | Han 2013 Results, "proportional residual error was chosen as the base PK model" |

The discussion paragraph below Table 3 provides two structural-model
sanity values that the encoded equations must reproduce:

* "When a patient was in hypermetabolic status with a CLCR of 120, the
  typical value of CL was expected to be 1.75 liters/h (sum of theta1,
  theta4, and theta7)." At CRRT = 0, CLCR = 120, TBI = 1, SEPS = 0:
  `CL = 0.693 + 0.557*1 + 0.504*1 + 0 = 1.754 L/h`. Matches.
* "For a patient within 30 days after injury, weighing 65 kg, the typical
  value of V was estimated to be 59.9 liters (sum of theta5 and theta8)
  when the patient had no edema." At WT = 65, EDEM = 0, TBI = 1:
  `V = (65/65)*50.3 + 0 + 9.61 = 59.91 L`. Matches.

# Virtual cohort

Original observed concentrations are not publicly available. The cohort
below reproduces the steady-state VPC design described in Han 2013
Methods (1,000 virtual patients with the covariate composition of the
original data set; concentrations simulated at the seven sampling time
points 3, 5, 9, 24, 27, 48, and 51 h after the start of infusion). The
vignette uses n = 200 per dose regimen (reduced from 1,000 to stay
inside the 5-minute pkgdown render budget; the median, 5th, and 95th
percentiles remain stable). The covariate composition is sampled to match
the Han 2013 Table 1 marginal distributions: CRRT 25%, sepsis 48%, edema
33%, recent-postburn (< 30 days) approximately 65% (consistent with mean
postburn time 23 days and majority of subjects in the acute hypermetabolic
window), weight log-normally distributed around mean 65.6 kg, and CLCR
log-normally distributed around mean 123.5 mL/min (truncated at the Table 1
range 21.6-282.7).

```{r cohort}
set.seed(20260609)

n_per_dose <- 200L
dose_levels <- c(100, 200, 400)    # mg q24h
infusion_min <- 25                 # minutes (Han 2013 Methods: 10-40 min, midpoint 25 min)
infusion_h <- infusion_min / 60
n_doses    <- 7L                   # 7 days q24h to reach steady state
tau_h      <- 24

# Sampling schedule matches Han 2013 Methods: 0, 3, 5, 9, 24, 27, 48, 51 h
# relative to the start of an arbitrary dosing interval at steady state.
# We pick the final 48 h window (interval (n_doses-1) and (n_doses)) so the
# trough at 48 h lands cleanly between two doses for SS interval comparison.
last_dose_time <- (n_doses - 1) * tau_h                 # 144 h
ss_window_start <- last_dose_time - tau_h               # 120 h (start of penultimate interval)
han_sample_offsets <- c(0, 3, 5, 9, 24, 27, 48, 51)     # h within penultimate interval
ss_sample_times <- ss_window_start + han_sample_offsets

# A denser grid for visual VPC overlay (covers two full intervals after
# steady state).
plot_sample_times <- seq(ss_window_start, ss_window_start + 51, by = 0.5)

sample_times <- sort(unique(c(ss_sample_times, plot_sample_times)))

# Helper draws one virtual cohort matched to Han 2013 Table 1 marginals.
draw_covariates <- function(n, id_offset = 0L) {
  # WT: truncated log-normal around the Table 1 mean 65.6 kg
  wt <- pmin(pmax(rlnorm(n, log(65.6) - 0.5 * 0.15^2, 0.15), 40), 90)
  # CLCR: truncated log-normal around the Table 1 mean 123.5 mL/min
  crcl <- pmin(pmax(rlnorm(n, log(123.5) - 0.5 * 0.45^2, 0.45),
                    21.6), 282.7)
  # Binary covariates per Table 1 marginal prevalences
  sepsis <- rbinom(n, 1, 29 / 60)
  edema  <- rbinom(n, 1, 20 / 60)
  crrt   <- rbinom(n, 1, 15 / 60)
  burn_recent <- rbinom(n, 1, 0.65)  # see Assumptions section
  data.frame(
    id              = id_offset + seq_len(n),
    WT              = wt,
    CRCL            = crcl,
    CRRT_STATUS     = crrt,
    DIS_SEPSIS      = sepsis,
    DIS_EDEMA       = edema,
    DIS_BURN_RECENT = burn_recent
  )
}

make_cohort <- function(dose_mg, id_offset) {
  covs <- draw_covariates(n = n_per_dose, id_offset = id_offset)
  dose_rows <- data.frame(
    id   = rep(covs$id, each = n_doses),
    time = rep(seq(0, by = tau_h, length.out = n_doses), times = n_per_dose),
    evid = 1L,
    amt  = dose_mg,
    rate = dose_mg / infusion_h,
    dv   = NA_real_
  )
  obs_rows <- data.frame(
    id   = rep(covs$id, each = length(sample_times)),
    time = rep(sample_times, times = n_per_dose),
    evid = 0L,
    amt  = NA_real_,
    rate = NA_real_,
    dv   = NA_real_
  )
  out <- rbind(dose_rows, obs_rows)
  out <- merge(out, covs, by = "id", all.x = TRUE)
  out$cohort  <- sprintf("%s mg q24h", dose_mg)
  out$dose_mg <- dose_mg
  out[order(out$id, out$time, -out$evid), ]
}

events_list <- vector("list", length(dose_levels))
for (i in seq_along(dose_levels)) {
  events_list[[i]] <- make_cohort(
    dose_mg   = dose_levels[i],
    id_offset = (i - 1L) * n_per_dose
  )
}
events <- dplyr::bind_rows(events_list)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod <- readModelDb("Han_2013_fluconazole")

sim <- rxode2::rxSolve(
  object = mod, events = events,
  keep   = c("cohort", "dose_mg", "WT", "CRCL", "CRRT_STATUS",
             "DIS_SEPSIS", "DIS_EDEMA", "DIS_BURN_RECENT")
) |>
  as.data.frame()
```

# Replicate published figures

## Figure 2 -- steady-state VPC by daily dose regimen

Han 2013 Figure 2 plots simulated 90% confidence intervals (5th-95th
percentiles) of fluconazole plasma concentrations grouped by the 100,
200, and 400 mg/day dosing regimens, with concentrations on a log scale.
The figure shows a single 48-h steady-state window. The vignette
reproduces the per-dose VPC ribbons and median curve at the matching
sampling time points.

```{r figure-2}
ss_vpc <- sim |>
  dplyr::filter(time %in% plot_sample_times) |>
  dplyr::mutate(time_h_in_ss_window = time - ss_window_start)

ss_summary <- ss_vpc |>
  dplyr::group_by(dose_mg, cohort, time_h_in_ss_window) |>
  dplyr::summarise(
    p05 = quantile(Cc, 0.05, na.rm = TRUE),
    p50 = quantile(Cc, 0.50, na.rm = TRUE),
    p95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    cohort = factor(cohort,
                    levels = sprintf("%s mg q24h", dose_levels))
  )

ggplot(ss_summary,
       aes(time_h_in_ss_window, p50, group = cohort)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "steelblue", alpha = 0.25) +
  geom_line(linewidth = 0.7, colour = "steelblue") +
  facet_wrap(~ cohort) +
  scale_y_log10() +
  labs(
    x        = "Time within steady-state window (h)",
    y        = "Fluconazole plasma concentration (ug/mL, log scale)",
    title    = "Figure 2 -- steady-state VPC by daily dose",
    subtitle = paste0("Replicates Han 2013 Figure 2 (100, 200, 400 mg/day); ",
                      "n = ", n_per_dose, " per dose; ",
                      "median (line) and 5th-95th percentile (ribbon).")
  ) +
  theme_minimal()
```

## Figure 3 -- probability of fAUC/MIC target attainment

Han 2013 Figure 3 plots the probability of target attainment (PTA) for
fAUC/MIC targets of 25 (upper panels) and 50 (lower panels) versus MIC
across patient subtypes that affect fluconazole CL (CRRT, sepsis, and
burn-recency status). The fAUC is computed as 0.88 * AUC_ss because
fluconazole is approximately 12% protein bound (Han 2013 Methods, citing
EUCAST 2007 reference 14); AUC_ss = daily dose / CL because fluconazole
PK is linear in the dose range studied. The replication below tabulates
PTA at MIC = 0.5, 1, and 2 ug/mL for 200 and 400 mg/day at three
patient profiles: CRRT, no-CRRT/sepsis/recent burn, and
no-CRRT/no-sepsis/resolved (no recent burn). All profiles use the median
CLCR (123.5 mL/min) and median WT (65 kg).

```{r figure-3}
# Per Han 2013 simulation: AUC_ss = daily_dose / CL at the typical-value level
# is shifted by IIV via the standard exp(eta) on CL. For PTA simulation,
# match the paper's approach: draw CL from the typical structural model under
# each patient profile and compute AUC = daily_dose / CL_i for each draw.

draw_cl <- function(n, profile, dose_mg) {
  # profile is a list with WT, CRCL, CRRT_STATUS, DIS_SEPSIS, DIS_EDEMA, DIS_BURN_RECENT
  if (profile$CRRT_STATUS == 1L) {
    eta <- rnorm(n, 0, sqrt(0.02559))
    cl_typ <- 1.85
  } else {
    eta <- rnorm(n, 0, sqrt(0.12000))
    cl_typ <- 0.693 +
              0.557 * (profile$CRCL / 120) +
              0.504 * profile$DIS_BURN_RECENT +
              (-0.369) * profile$DIS_SEPSIS
  }
  cl_typ * exp(eta)
}

pta_one <- function(dose_mg, profile, target, mics, n_draws = 10000L) {
  cl <- draw_cl(n_draws, profile, dose_mg)
  auc <- dose_mg / cl                          # ug*h/mL = mg*h/L (since 1 mg/L = 1 ug/mL)
  fauc <- 0.88 * auc                           # 12% protein binding -> 88% free
  sapply(mics, function(m) mean(fauc / m >= target))
}

profiles <- list(
  "CRRT (1g equivalent)" = list(
    WT = 65, CRCL = 0, CRRT_STATUS = 1L,
    DIS_SEPSIS = 0L, DIS_EDEMA = 0L, DIS_BURN_RECENT = 0L
  ),
  "non-CRRT, septic, recent burn" = list(
    WT = 65, CRCL = 123.5, CRRT_STATUS = 0L,
    DIS_SEPSIS = 1L, DIS_EDEMA = 0L, DIS_BURN_RECENT = 1L
  ),
  "non-CRRT, no sepsis, resolved" = list(
    WT = 65, CRCL = 123.5, CRRT_STATUS = 0L,
    DIS_SEPSIS = 0L, DIS_EDEMA = 0L, DIS_BURN_RECENT = 0L
  )
)

mic_grid <- c(0.5, 1, 2)
pta_grid <- expand.grid(
  profile = names(profiles),
  dose_mg = c(200, 400),
  target  = c(25, 50),
  stringsAsFactors = FALSE
)

pta_rows <- vector("list", nrow(pta_grid))
for (i in seq_len(nrow(pta_grid))) {
  row <- pta_grid[i, ]
  vals <- pta_one(
    dose_mg = row$dose_mg,
    profile = profiles[[row$profile]],
    target  = row$target,
    mics    = mic_grid
  )
  pta_rows[[i]] <- data.frame(
    profile = row$profile,
    dose_mg = row$dose_mg,
    target  = row$target,
    mic     = mic_grid,
    pta     = round(vals, 3)
  )
}
pta_long <- dplyr::bind_rows(pta_rows)

pta_wide <- pta_long |>
  tidyr::pivot_wider(
    names_from  = mic,
    values_from = pta,
    names_glue  = "PTA at MIC = {mic} ug/mL"
  ) |>
  dplyr::arrange(profile, target, dose_mg)

knitr::kable(
  pta_wide,
  caption = paste0("Probability of fAUC/MIC target attainment (PTA) ",
                   "for fluconazole 200 and 400 mg/day across three ",
                   "patient profiles, computed from 10,000 Monte-Carlo ",
                   "draws of CL under the Han 2013 covariate model. ",
                   "Compare against Han 2013 Figure 3.")
)
```

Han 2013 Figure 3 reports near-100 % PTA for 400 mg/day at a target of
25 across all profiles when MIC <= 2 ug/mL, and shows the CRRT / recent-
burn profiles producing markedly lower PTA at the higher target (50)
because their CL is higher. The simulated table reproduces this
qualitative pattern: the recent-burn / non-CRRT / septic profile and the
CRRT profile have lower PTA than the resolved-phase reference at the
stricter target of 50.

# PKNCA validation

PKNCA computes the steady-state AUC0-tau on the penultimate dosing
interval (`ss_window_start` to `ss_window_start + tau_h`) across the
three dose regimens. The treatment grouping in the formula is the
`cohort` label so per-dose Cmax, Cmin, Cavg, and AUC0-tau are summarised
together.

```{r pknca}
pk_conc <- sim |>
  dplyr::filter(time >= ss_window_start, time <= ss_window_start + tau_h,
                !is.na(Cc), Cc > 0) |>
  dplyr::select(id, time, Cc, cohort)

pk_dose <- events |>
  dplyr::filter(evid == 1L,
                time >= ss_window_start, time < ss_window_start + tau_h) |>
  dplyr::select(id, time, amt, cohort)

conc_obj <- PKNCA::PKNCAconc(
  data    = pk_conc,
  formula = Cc ~ time | cohort + id,
  concu   = "ug/mL",
  timeu   = "hr"
)
dose_obj <- PKNCA::PKNCAdose(
  data    = pk_dose,
  formula = amt ~ time | cohort + id,
  doseu   = "mg"
)

intervals <- data.frame(
  start   = ss_window_start,
  end     = ss_window_start + tau_h,
  cmax    = TRUE,
  tmax    = TRUE,
  cmin    = TRUE,
  cav     = TRUE,
  auclast = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- suppressWarnings(PKNCA::pk.nca(nca_data))

knitr::kable(
  summary(nca_res),
  caption = paste0("Steady-state NCA parameters (Cmax,ss, Cmin,ss, ",
                   "Cavg,ss, AUC0-tau,ss) by dose regimen at ",
                   "the penultimate dosing interval; n = ",
                   n_per_dose, " per regimen.")
)
```

## Comparison against derived AUC_ss

Han 2013 did not publish an NCA table directly. The Methods section,
however, states that the simulated 24-h steady-state AUC was computed as
`daily_dose / CL` because fluconazole shows linear PK in the dose range
studied. For the resolved-phase typical patient (CRRT = 0, CLCR = 120,
TBI = 0, SEPS = 0), `CL = 0.693 + 0.557 = 1.25 L/h` (Han 2013 Discussion
explicit value), so:

| Daily dose | Expected AUC_ss (mg*h/L = ug*h/mL) |
|------------|-------------------------------------|
| 100 mg     | 100 / 1.25 = 80                     |
| 200 mg     | 200 / 1.25 = 160                    |
| 400 mg     | 400 / 1.25 = 320                    |

These expected AUC_ss values are the typical-value reference; the
simulated PKNCA AUC0-tau median should be close to the expected value at
the resolved-phase typical patient. The simulated medians above include
the full cohort with CRRT (CL = 1.85, AUC reduced by 1.25/1.85), recent
burn (CL increased), and sepsis (CL decreased) variation, so the cohort
median AUC0-tau will be slightly below the resolved-phase typical-patient
value of 80 / 160 / 320 ug*h/mL.

# Assumptions and deviations

- **`theta2` (V intercept) dropped per Han 2013 Table 3 footnote b.** The
  Table 3 row for `theta2` records "NE" (not estimated) with the footnote
  "In the final model, when the covariates were fully included, the
  slope of the proportional relationship (theta2) was not successfully
  estimated." The packaged model encodes the V equation as
  `V = (WT/65)*exp(lvc) + e_dis_edema_vc*EDEM + e_dis_burn_recent_vc*TBI`
  with `theta2 = 0`, consistent with the Han 2013 Discussion line "For a
  patient within 30 days after injury, weighing 65 kg, the typical value
  of V was estimated to be 59.9 liters (sum of theta5 and theta8)" --
  no theta2 contribution is included in the published reference value.

- **CLCR normalisation factor 120 mL/min from Han 2013 Table 3 parameter
  description.** Han 2013 Table 2 writes the model-development equations
  as `CL_RENAL = (theta1 + CLCR * theta4) * exp(eta1)` (CLCR as bare
  mL/min), but Han 2013 Table 3 describes `theta4` as the "Proportionality
  constant between CL and CLCR/120" and Han 2013 Discussion reports that
  at CLCR = 120 the CL contribution from this term equals `theta4` = 0.557
  L/h (which is consistent only if `(CLCR/120) * theta4` is used, not
  `CLCR * theta4`). The packaged model uses the Table 3 / Discussion
  form `(CLCR / 120) * theta4`; the Table 2 form appears to be a
  typographical contraction of the normalised version.

- **CRCL stored under the canonical `CRCL` column despite NOT being
  BSA-normalized.** Han 2013 uses raw Cockcroft-Gault CrCL (mL/min, not
  BSA-normalized). Following the precedent of `Delattre_2010_amikacin.R`
  and `Shekar_2014_meropenem.R`, the model stores the source `CLCR`
  column under `CRCL`, with the raw / non-BSA-normalized status documented
  in the per-model `covariateData[[CRCL]]$units` and `notes` fields. Do
  not compare the magnitude of `e_crcl_cl = 0.557` against BSA-normalized
  reference values listed in the canonical entry -- the units differ.

- **Three new canonical covariates ratified alongside this extraction.**
  `DIS_SEPSIS`, `DIS_EDEMA`, and `DIS_BURN_RECENT` are added to
  `inst/references/covariate-columns.md`. The Han 2013 source column
  names `SEPS`, `EDEM`, and `TBI` are recorded as aliases in each
  respective entry. The `TBI` alias carries an explicit note that this
  Han 2013 column does NOT denote traumatic brain injury -- it is a
  recoded binary indicator at the 30-day postburn threshold.

- **Conditional IIV split by CRRT status; `etalcl_rrt` declared as
  paper-specific.** Han 2013 reports two distinct CL BSV estimates
  (omega1^2 = 35.7 %CV for the non-RRT arm, omega3^2 = 16.1 %CV for the
  RRT arm). The packaged model encodes this by pairing each arm with its
  own multiplicative eta (`etalcl` for non-RRT, `etalcl_rrt` for RRT).
  `etalcl_rrt` is declared in `paper_specific_etas` at the top of the
  model function because the conditional-IIV split-by-CRRT structure
  does not match the 1-to-1 eta-to-lparam pairing rule enforced by
  `checkModelConventions()`. The non-applicable arm's eta evaluates with
  the indicator-zeroed multiplier in `model()`, so each subject's
  per-time-point CL only carries the IIV from the arm matching their
  CRRT status.

- **Additive covariate model implemented in linear space.** Han 2013
  reports `CL = theta1 + ...` and `V = theta5*WT/65 + ...` as additive
  formulas, not multiplicative. The packaged model preserves the
  additive structure by computing the linear sum first and then
  multiplying by `exp(eta)`; the structural log-transformed `lcl`,
  `lcl_rrt`, and `lvc` are back-transformed via `exp(...)` inside
  `model()` so only positive intercepts enter the additive sum. The
  additive covariate coefficients (`e_crcl_cl`, `e_dis_burn_recent_cl`,
  `e_dis_sepsis_cl`, `e_dis_edema_vc`, `e_dis_burn_recent_vc`) are kept
  in linear space because they can take negative values (the sepsis
  coefficient is -0.369 L/h).

- **`propSd <- sqrt(0.111) = 0.3332` not `0.111`.** Han 2013 Table 3
  reports the proportional residual error as `sigma1^2 = 0.111`, the
  NONMEM `$SIGMA` value (variance on the proportional residual scale,
  not SD). For nlmixr2's `Cc ~ prop(propSd)` the `propSd` argument is
  the SD of the proportional residual, so `propSd = sqrt(0.111) = 0.3332`.

- **Recent-postburn (`DIS_BURN_RECENT = 1`) prevalence in the virtual
  cohort set to ~65 %.** Han 2013 Table 1 reports the mean time from burn
  injury was 23 days (range 7-88), which places the majority of the
  cohort in the acute hypermetabolic window (within 30 days), but the
  paper does not tabulate the exact `TBI` prevalence. The vignette uses
  65 % as a reasonable approximation (the mean of 23 days suggests well
  over half the cohort had time from burn injury < 30 days); changing
  this rate affects only the marginal cohort statistics in Figure 2,
  not the per-subject typical-value predictions, and not the PTA table
  (which fixes the patient profile rather than sampling it).

- **Infusion duration set to 25 min (midpoint of Han 2013 Methods).**
  Han 2013 Methods reports fluconazole was "intravenously infused with
  100 to 400 mg of fluconazole for 10 to 40 min every 24 h"; the
  vignette uses 25 min as the midpoint. The exact infusion duration
  affects the very early time-course (Cmax shape during infusion) but
  has negligible effect on AUC and on Cmin / Cavg / Ctau at steady
  state, which are the quantities the validation focuses on.

- **Steady state defined at 7 days of q24h dosing.** Han 2013 sampled
  patients who had received 5-17 days (mean 6.5 days) of fluconazole
  therapy at the start of PK sampling. The vignette uses 7 doses (7 days)
  before sampling the penultimate interval at days 5-7; with the
  typical-patient terminal half-life of ln(2) * 50.3 / 1.25 = 27.9 h,
  steady state is reached within 5-6 days, so 7 days is sufficient.

- **Race / ethnicity not modeled.** Han 2013 enrolled a single-centre
  Korean cohort. The paper does not test race as a covariate; the model
  does not include a race effect.

- **No additive residual error in the final model.** Han 2013 Methods
  describes the candidate residual-error structure as a combined additive
  plus proportional model. Han 2013 Results section ("PK model")
  states: "A one-compartment first-order elimination model with
  proportional residual error was chosen as the base PK model." Only the
  proportional component is reported in Table 3 (sigma1^2 = 0.111); no
  additive sigma is reported. The packaged model uses
  `Cc ~ prop(propSd)` (no additive component).
