---
title: "Peg-IFN-alpha-induced HBsAg loss in chronic hepatitis B (Hanan 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Peg-IFN-alpha-induced HBsAg loss in chronic hepatitis B (Hanan 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7.5, fig.height = 5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
rxode2::rxSetSeed(20260829)
```

```{r load-models, include = FALSE}
ui_eot  <- rxode2::rxode(readModelDb("Hanan_2026_peginterferon_alfa_eot_mbma"))
ui_24wk <- rxode2::rxode(readModelDb("Hanan_2026_peginterferon_alfa_24wk_mbma"))

# Pull a fixed-effect estimate out of an rxUi by name, failing loudly if the
# name is absent or ambiguous. Every published-value comparison below reads the
# encoded value through this helper rather than restating a literal, so the
# checks test the model file rather than this vignette's own arithmetic.
getpar <- function(ui, nm) {
  v <- ui$iniDf$est[ui$iniDf$name == nm]
  if (length(v) != 1L) stop("no unique ini() entry named '", nm, "'")
  v
}
```

# Model and source

* Citation: `r ui_eot$reference`
* Article: <https://doi.org/10.1002/psp4.70164> (Open Access; PMC12823784)
* Supplement (Data S1): PRISMA flow diagram (Figure S1) and the PICOS
  inclusion / exclusion table (Table S1). It contains **no parameters**;
  it contributed only population metadata to this extraction.
* Errata: none. Crossref `update-to` and `updated-by` are both null.

This paper contributes **two** model files, because the authors fit their two
efficacy endpoints independently (Section 2.3) on different analysis sets with
different retained covariates:

| Model file | Endpoint | Analysis set | Retained covariates |
|---|---|---|---|
| `Hanan_2026_peginterferon_alfa_eot_mbma` | HBsAg loss at end of treatment (EOT) | 45 studies / 52 strata / 83 study-strata-arms / 11,493 participants | duration, baseline HBsAg, **HBeAg status** |
| `Hanan_2026_peginterferon_alfa_24wk_mbma` | HBsAg loss 24 weeks after Peg-IFN-alpha cessation (functional-cure surrogate) | 28 studies / 35 strata / 58 study-strata-arms / 4,267 participants | duration, baseline HBsAg |

Baseline HBeAg status is the one structural difference between them: it is
retained at end of treatment (OR 2.37) and screened-but-not-retained at 24
weeks. The 24-week model therefore predicts the same probability for
HBeAg-positive and HBeAg-negative arms, and the EOT coefficient must **not** be
carried over to it.

## Why this is an extraction and not a systematic-review skip

Page 1 carries a "SYSTEMIC REVIEW" banner and the Methods describe a PRISMA
literature search of Embase / MEDLINE / Cochrane (January 2000 to July 2022),
so the systematic-review skip pathway applies to the *data collection* step
only. The literature search is how the authors assembled their dataset; the
model itself is original, fit by the authors in R 4.2.1 with `nlme`, and Table 2
is their own parameter table (logit-scale estimates, %RSE, 95% prediction
intervals, p-values, between-trial variance) rather than a catalogue of other
authors' models.

# Population

```{r population, echo = FALSE}
pop <- ui_eot$population
data.frame(
  Field = c("Species", "Disease state", "Participants (EOT / 24-week)",
            "Studies (EOT / 24-week)", "Study-strata-arms (EOT / 24-week)",
            "Age", "Regimens"),
  Value = c(
    pop$species,
    pop$disease_state,
    paste(format(pop$n_subjects, big.mark = ","), "/",
          format(ui_24wk$population$n_subjects, big.mark = ",")),
    paste(pop$n_studies, "/", ui_24wk$population$n_studies),
    paste(pop$n_study_strata_arms, "/", ui_24wk$population$n_study_strata_arms),
    pop$age_range,
    pop$dose_range
  )
) |>
  knitr::kable(caption = "Population metadata, read from the model files' `population` lists.")
```

The observation unit is the **study-strata-arm**: a treatment arm within a
study, optionally stratified by HBeAg status, baseline HBsAg or baseline HBV
DNA. The data are aggregate published results, not individual patient data, so
every simulation in this vignette produces a **study-arm mean proportion** and
none of it is an individual-patient prediction. Analysis-set exclusions (Section
2.2 and Table S1) mean the models do not describe pregnant or post-partum women,
HBV/HIV coinfection, post-liver-transplant or immunosuppressed participants.

The same metadata is available programmatically via
`readModelDb("Hanan_2026_peginterferon_alfa_eot_mbma")()$population`.

# Model structure

Both models use the Section 2.3.3 final covariate form. For study `i`, stratum
`s` and treatment arm `a`:

$$
p_{isa} = \mathrm{logit}^{-1}\!\left( \mathrm{drug.eff}_d
   + \sum_m \beta_m\, \theta_{m,isa} + \eta_{isa} \right),
\qquad \eta_{isa} \sim N(0, \sigma^2_\eta)
$$

`drug.eff_d` is a regimen-specific intercept (Section 2.3.2) selected by the
combination-therapy indicator `CONMED_NUC`, so the model file carries two
intercepts rather than one intercept plus a contrast coefficient -- that is the
form the authors estimated and reported. Continuous covariates enter centered at
the across-arm median (48 weeks; 3 log10 IU/mL).

There are **no ODE states, no dose events and no time axis**. Both models are
cross-sectional: Section 2.2 states that two cross-sectional models were
developed rather than a longitudinal trajectory model because too few trials
reported both endpoints or intermediate follow-up. The `time` column in the
event tables below is therefore inert, and observation rows carry
`cmt = NA_character_` because there is no compartment to align to.

`eta_study` is a **between-study-strata-arm** random effect on the logit scale.
It is not a popPK between-subject variance: it describes how far a new
*study-arm's* underlying log-odds sits from the typical arm, and says nothing
about individual patients within an arm. It is declared
`paper_specific_etas <- c("eta_study")` rather than being relabelled into the
`eta<transformed-parameter>` popPK convention.

Section 2.3.1 fixes the residual variance to 1 and applies the binomial variance
function `w_isa = sqrt(p (1 - p) / N_isa)`, so there is no estimated
residual-error parameter. `addSd` is `fixed(0.001)` purely for nlmixr2 endpoint
compatibility (the `Goteti_2024_SLE_mbma` device), and `model()` exposes
`w_binom_unit = sqrt(p (1 - p))` so a user recovers the paper's weight as
`w_binom_unit / sqrt(N_isa)` for an arm of size `N_isa`.

# Source trace

Per-parameter provenance is recorded as an in-file comment beside each `ini()`
entry in `inst/modeldb/specificDrugs/Hanan_2026_peginterferon_alfa_eot_mbma.R`
and `..._24wk_mbma.R`. Collected here for review; every value is from Hanan 2026
Table 2, "Estimate (logit)" column.

| Parameter | EOT model | 24-week model | Source |
|---|---|---|---|
| `logitp_mono` | -2.37 | -2.08 | Table 2, "Drug effect (Peg-IFNa)" |
| `logitp_comb` | -2.02 | -1.81 | Table 2, "Drug effect (Peg-IFNa +NA)" |
| `e_t_pegifn_logitp` | 0.0133 | 0.0303 | Table 2, "Peg-IFNa treatment duration (per week, ref.=48 weeks)" |
| `e_hbsag_bl_log10_logitp` | -1.57 | -1.58 | Table 2, "Baseline HBsAg (per log10 IU/mL, ref.=3)" |
| `e_hbeag_pos_logitp` | 0.863 | not retained | Table 2, "HBeAg status (ref = negative)" |
| `eta_study` (variance) | 0.484 | 0.572 | Table 2, "Between-trial variance" |
| `addSd` | `fixed(0.001)` | `fixed(0.001)` | not a paper parameter; see Section 2.3.1 note above |
| logit-linear structure | n/a | n/a | Section 2.3.1 / 2.3.2 / 2.3.3 |
| binomial variance function | n/a | n/a | Section 2.3.1 |

Covariate columns use four canonical names, all registered in
`inst/references/covariate-columns.md` as part of this extraction:
`HBSAG_BL_LOG10`, `HBEAG_POS`, `T_PEGIFN` and `CONMED_NUC`.

# Virtual cohort

A "cohort" here is a set of **study-arm scenarios**, not a set of patients: one
row per hypothetical trial arm, each with its regimen, planned Peg-IFN-alpha
duration, baseline HBsAg and HBeAg stratum.

```{r cohort-builder}
# One row per study-arm scenario. `time` is inert (the models are
# cross-sectional) and `cmt` is NA because neither model declares an ODE state.
make_arms <- function(arms) {
  arms |>
    mutate(
      id   = row_number(),
      time = 0,
      amt  = 0,
      evid = 0L,
      cmt  = NA_character_
    )
}

# Typical-value (zeroRe) versions: eta_study is set to zero so the prediction is
# the typical study-arm rather than a draw from the between-trial distribution.
mod_eot_typ  <- rxode2::zeroRe(readModelDb("Hanan_2026_peginterferon_alfa_eot_mbma"))
mod_24wk_typ <- rxode2::zeroRe(readModelDb("Hanan_2026_peginterferon_alfa_24wk_mbma"))

solve_arms <- function(mod, arms, keep = character()) {
  rxode2::rxSolve(mod, events = make_arms(arms), keep = keep) |>
    as.data.frame()
}
```

# Check 1 -- Table 2 reference population back-transforms

Table 2's "Transformed estimate" column reports each drug effect as a
probability in the reference population: HBeAg-negative, baseline HBsAg
3 log10 IU/mL, 48 weeks of Peg-IFN-alpha. The two additional EOT rates in the
Table 2 Interpretation column give the HBeAg-positive reference population.

```{r check1-reference-rates}
ref_arms <- tibble::tibble(
  scenario       = c("Peg-IFNa monotherapy", "Peg-IFNa + NA",
                     "Peg-IFNa monotherapy", "Peg-IFNa + NA"),
  HBEAG_POS      = c(0, 0, 1, 1),
  CONMED_NUC     = c(0, 1, 0, 1),
  T_PEGIFN       = 48,
  HBSAG_BL_LOG10 = 3
)

ref_eot <- solve_arms(mod_eot_typ, ref_arms,
                      keep = c("scenario", "HBEAG_POS", "CONMED_NUC"))
# The 24-week model does not use HBEAG_POS, so it needs only the two regimens.
ref_24wk <- solve_arms(mod_24wk_typ, ref_arms[1:2, ],
                       keep = c("scenario", "CONMED_NUC"))

check1 <- tibble::tibble(
  quantity = c(
    "EOT, monotherapy, HBeAg-negative",
    "EOT, combination, HBeAg-negative",
    "EOT, monotherapy, HBeAg-positive",
    "EOT, combination, HBeAg-positive",
    "24-week, monotherapy",
    "24-week, combination"
  ),
  model_pct     = 100 * c(ref_eot$p_hbsag_loss, ref_24wk$p_hbsag_loss),
  published_pct = c(8.5, 11.7, 18.1, 23.8, 11.1, 14.1),
  source = c(rep("Table 2 transformed estimate / Interpretation", 4),
             rep("Table 2 transformed estimate", 2))
) |>
  mutate(difference_pp = model_pct - published_pct)

check1 |>
  dplyr::rename(
    "Quantity"             = quantity,
    "Model (%)"            = model_pct,
    "Published (%)"        = published_pct,
    "Difference (pp)"      = difference_pp,
    "Source"               = source
  ) |>
  knitr::kable(digits = 2, caption = "Table 2 reference-population HBsAg-loss rates reproduced through rxode2.")

# Table 2 prints these to 3 significant figures, so the rounding tolerance is
# 0.05 pp on the two-decimal rows and 0.1 pp on the one-decimal rows.
stopifnot(all(abs(check1$difference_pp) < 0.13))
```

The 23.92% vs 23.8% row carries the largest residual (`r sprintf("%.2f", max(abs(check1$difference_pp)))` pp) and is
pure rounding: the Interpretation column quotes the HBeAg-positive combination
rate to three significant figures.

# Check 2 -- odds ratios recovered by finite contrast

Exponentiating an `ini()` value would only restate the input. Instead each odds
ratio is recovered from the **solved model** as the ratio of odds at
`covariate + 1` to odds at `covariate`, which tests the `model()` wiring
(centering, sign, which intercept the indicator selects).

```{r check2-odds-ratios}
odds <- function(p) p / (1 - p)

# Contrast a one-unit change in `cov` against a baseline arm.
or_from_model <- function(mod, base_arm, cov, delta = 1) {
  bumped <- base_arm
  bumped[[cov]] <- bumped[[cov]] + delta
  s <- solve_arms(mod, dplyr::bind_rows(base_arm, bumped))
  odds(s$p_hbsag_loss[2]) / odds(s$p_hbsag_loss[1])
}

base_arm <- tibble::tibble(
  HBEAG_POS = 0, CONMED_NUC = 0, T_PEGIFN = 48, HBSAG_BL_LOG10 = 3
)

check2 <- tibble::tibble(
  quantity = c(
    "EOT, OR per additional treatment week",
    "EOT, OR per log10 IU/mL baseline HBsAg",
    "EOT, OR for HBeAg-positive vs negative",
    "24-week, OR per additional treatment week",
    "24-week, OR per log10 IU/mL baseline HBsAg"
  ),
  model_or = c(
    or_from_model(mod_eot_typ,  base_arm, "T_PEGIFN"),
    or_from_model(mod_eot_typ,  base_arm, "HBSAG_BL_LOG10"),
    or_from_model(mod_eot_typ,  base_arm, "HBEAG_POS"),
    or_from_model(mod_24wk_typ, base_arm, "T_PEGIFN"),
    or_from_model(mod_24wk_typ, base_arm, "HBSAG_BL_LOG10")
  ),
  published_or = c(1.013, 0.21, 2.37, 1.03, 0.21)
)

check2 |>
  dplyr::rename(
    "Quantity"        = quantity,
    "Model OR"        = model_or,
    "Published OR"    = published_or
  ) |>
  knitr::kable(digits = 4, caption = "Table 2 odds ratios recovered by finite contrast on the solved model.")

# Each contrast must land within the rounding of the value Table 2 prints. Note
# the two duration rows are printed to DIFFERENT precision -- "OR 1.013" at EOT
# but "OR 1.03" at 24 weeks -- so the EOT row gets the tighter bound.
stopifnot(
  abs(check2$model_or - check2$published_or) <
    c(0.0005, 0.005, 0.005, 0.005, 0.005)
)
```

The baseline-HBsAg odds ratio also reproduces the Results Section 3.3 sentence
"for each log10 unit increase in baseline HBsAg, the odds of achieving HBsAg
loss decreased by 79%": the EOT model gives a
`r sprintf("%.1f", 100 * (1 - check2$model_or[2]))`% reduction and the 24-week
model `r sprintf("%.1f", 100 * (1 - check2$model_or[5]))`%.

# Check 3 -- the between-trial variance is a variance, not a standard deviation

Table 2 reports a 95% prediction interval alongside each drug effect, and
Section 3.4 states these "incorporate both parameter uncertainty and
between-trial variability". On the logit scale those two sources are additive,
so the interval is reconstructable in closed form as

`logit^-1( estimate +/- 1.96 * sqrt(between-trial variance + SE^2) )`

with `SE = %RSE/100 * |estimate|` from the Table 2 %RSE column. This is the
sharpest available test that `eta_study` was encoded as a **variance** and not
as a standard deviation or a CV: reading 0.484 as an SD would put the EOT
monotherapy upper bound near 20% instead of the published 27.7%.

```{r check3-prediction-intervals}
expit <- function(x) 1 / (1 + exp(-x))

pi_recon <- function(est, rse_pct, omega_var) {
  se <- rse_pct / 100 * abs(est)
  sd_total <- sqrt(omega_var + se^2)
  100 * expit(est + c(-1, 1) * qnorm(0.975) * sd_total)
}

omega_eot  <- getpar(ui_eot,  "eta_study")
omega_24wk <- getpar(ui_24wk, "eta_study")

check3 <- tibble::tibble(
  quantity = c("EOT, monotherapy", "EOT, combination",
               "24-week, monotherapy", "24-week, combination"),
  est = c(getpar(ui_eot, "logitp_mono"), getpar(ui_eot, "logitp_comb"),
          getpar(ui_24wk, "logitp_mono"), getpar(ui_24wk, "logitp_comb")),
  # Table 2 %RSE column.
  rse_pct   = c(7.7, 10.3, 9.1, 11.9),
  omega_var = c(omega_eot, omega_eot, omega_24wk, omega_24wk),
  pub_lower = c(2.3, 3.1, 2.7, 3.4),
  pub_upper = c(27.7, 35.5, 36.6, 43.4)
) |>
  rowwise() |>
  mutate(
    recon_lower = pi_recon(est, rse_pct, omega_var)[1],
    recon_upper = pi_recon(est, rse_pct, omega_var)[2]
  ) |>
  ungroup()

check3 |>
  transmute(
    quantity,
    recon = sprintf("%.1f-%.1f", recon_lower, recon_upper),
    pub   = sprintf("%.1f-%.1f", pub_lower, pub_upper),
    max_diff_pp = pmax(abs(recon_lower - pub_lower), abs(recon_upper - pub_upper))
  ) |>
  dplyr::rename(
    "Quantity"                  = quantity,
    "Reconstructed 95% PI (%)"  = recon,
    "Published 95% PI (%)"      = pub,
    "Max difference (pp)"       = max_diff_pp
  ) |>
  knitr::kable(digits = 2, caption = "Table 2 95% prediction intervals reconstructed from the encoded between-trial variance plus the published %RSE.")

# The paper simulated these with 100,000 replicates, so a closed-form
# reconstruction should agree to well within a tenth of a percentage point.
stopifnot(
  abs(check3$recon_lower - check3$pub_lower) < 0.15,
  abs(check3$recon_upper - check3$pub_upper) < 0.15
)

# The discriminating counterfactual: if 0.484 were a standard deviation, the
# EOT monotherapy upper bound would be nowhere near the published 27.7%.
as_sd <- pi_recon(getpar(ui_eot, "logitp_mono"), 7.7, omega_eot^2)[2]
stopifnot(abs(as_sd - 27.7) > 5)
```

Reading the between-trial term as a standard deviation would give an upper bound
of `r sprintf("%.1f", as_sd)`% against a published 27.7%, so the variance
encoding is confirmed rather than assumed.

Sampling the same variance through rxode2 rather than in closed form gives a
consistent typical arm:

```{r check3-monte-carlo}
n_arms <- 200  # study-arm draws; the model has no individual-patient level
mc_arms <- ref_arms[1, ][rep(1, n_arms), ]
mc <- solve_arms(readModelDb("Hanan_2026_peginterferon_alfa_eot_mbma"), mc_arms)

mc_median_pct <- 100 * median(mc$p_hbsag_loss)
typical_pct   <- 100 * ref_eot$p_hbsag_loss[1]

# eta_study is symmetric on the logit scale, so the MEDIAN of the sampled arms
# is the typical-value prediction. Asserting on the median (not on the extremes
# of a random cohort) keeps this stable across rxode2 versions.
stopifnot(abs(mc_median_pct - typical_pct) < 1.5)
```

The median of `r n_arms` sampled study-arms is
`r sprintf("%.2f", mc_median_pct)`% against a typical value of
`r sprintf("%.2f", typical_pct)`%.

# Check 4 -- Table 3, the paper's 100,000-replicate simulation answer key

Table 3 simulates HBsAg loss for eight hypothetical phase 3 arms (four baseline
HBsAg criteria, two regimens) at a standard 48-week duration, reporting the mean
baseline HBsAg and the HBeAg-positive percentage of each arm.

Reproducing an arm rate needs one piece of care. The model is logit-linear in
`HBEAG_POS`, so an arm containing both strata has an aggregate rate equal to the
HBeAg-**weighted average of the two subgroup probabilities**, not the
probability evaluated at the mean indicator. The 24-week model does not retain
HBeAg, so this only affects the EOT column.

```{r check4-table3}
table3 <- tibble::tribble(
  ~regimen,        ~conmed_nuc, ~criteria, ~hbeag_pct, ~mean_bl, ~eot_pub, ~wk24_pub,
  "Peg-IFNa mono",           0,   "<= 5",         27,      3.5,      5.1,       5.4,
  "Peg-IFNa mono",           0,   "<= 4",         23,      3.2,      7.7,       8.4,
  "Peg-IFNa mono",           0,   "<= 3",         14,      2.6,     16.5,      19.1,
  "Peg-IFNa mono",           0,   "<= 2",          0,      1.9,     34.5,      41.6,
  "Peg-IFNa + NA",           1,   "<= 5",         31,      3.3,      7.1,       6.8,
  "Peg-IFNa + NA",           1,   "<= 4",         30,      3.2,     10.6,      10.4,
  "Peg-IFNa + NA",           1,   "<= 3",         20,      2.6,     21.9,      23.0,
  "Peg-IFNa + NA",           1,   "<= 2",          0,      1.9,     42.7,      47.5
)

# EOT: solve each row twice (HBeAg-negative and HBeAg-positive) and take the
# HBeAg-weighted average of the two probabilities.
eot_arms <- bind_rows(
  table3 |> transmute(row = row_number(), HBEAG_POS = 0, CONMED_NUC = conmed_nuc,
                      T_PEGIFN = 48, HBSAG_BL_LOG10 = mean_bl),
  table3 |> transmute(row = row_number(), HBEAG_POS = 1, CONMED_NUC = conmed_nuc,
                      T_PEGIFN = 48, HBSAG_BL_LOG10 = mean_bl)
)
eot_sol <- solve_arms(mod_eot_typ, eot_arms, keep = c("row", "HBEAG_POS"))

eot_model <- eot_sol |>
  select(row, HBEAG_POS, p = p_hbsag_loss) |>
  pivot_wider(names_from = HBEAG_POS, values_from = p,
              names_prefix = "hbeag_") |>
  left_join(table3 |> mutate(row = row_number()), by = "row") |>
  mutate(eot_model = 100 * (hbeag_pct / 100 * hbeag_1 +
                            (1 - hbeag_pct / 100) * hbeag_0))

# 24-week: HBeAg is not in the model, so one solve per row.
wk24_sol <- solve_arms(
  mod_24wk_typ,
  table3 |> transmute(row = row_number(), CONMED_NUC = conmed_nuc,
                      T_PEGIFN = 48, HBSAG_BL_LOG10 = mean_bl),
  keep = "row"
)

check4 <- eot_model |>
  select(row, regimen, criteria, hbeag_pct, mean_bl, eot_model, eot_pub, wk24_pub) |>
  left_join(wk24_sol |> transmute(row, wk24_model = 100 * p_hbsag_loss), by = "row") |>
  mutate(eot_diff = eot_model - eot_pub, wk24_diff = wk24_model - wk24_pub)

check4 |>
  transmute(
    regimen, criteria, hbeag_pct, mean_bl,
    eot_model, eot_pub, eot_diff, wk24_model, wk24_pub, wk24_diff
  ) |>
  dplyr::rename(
    "Treatment group"          = regimen,
    "HBsAg criteria (log10)"   = criteria,
    "HBeAg positive (%)"       = hbeag_pct,
    "Mean BL HBsAg (log10)"    = mean_bl,
    "EOT model (%)"            = eot_model,
    "EOT published (%)"        = eot_pub,
    "EOT diff (pp)"            = eot_diff,
    "24-wk model (%)"          = wk24_model,
    "24-wk published (%)"      = wk24_pub,
    "24-wk diff (pp)"          = wk24_diff
  ) |>
  knitr::kable(digits = 2, caption = "Hanan 2026 Table 3 reproduced. All rows at 48 weeks of Peg-IFN-alpha.")
```

Three separate assertions, because the rows differ in what they can prove.

```{r check4-assertions}
pure <- check4 |> filter(hbeag_pct == 0)   # no HBeAg mixture to approximate
mixed <- check4 |> filter(hbeag_pct > 0)
flagged <- check4 |> filter(regimen == "Peg-IFNa + NA", criteria == "<= 5")

# (a) The two 0%-HBeAg EOT rows have no mixture approximation in them, so they
#     must reproduce essentially exactly.
stopifnot(abs(pure$eot_diff) < 0.1)

# (b) The 24-week model never uses HBeAg, so the whole column is a clean test --
#     except the row flagged in the Errata below, whose printed baseline is
#     internally inconsistent with its own printed rate.
stopifnot(
  abs(check4$wk24_diff[check4$row != flagged$row]) < 0.8
)

# (c) The HBeAg-mixed EOT rows carry a known positive bias from assuming HBeAg
#     status and baseline HBsAg are independent within the arm (see below). The
#     bias must be positive and small on every row except the flagged one.
mixed_ok <- mixed |> filter(row != flagged$row)
stopifnot(mixed_ok$eot_diff > 0, mixed_ok$eot_diff < 1.6)
```

The residual bias on the HBeAg-mixed EOT rows is mechanistic, not a
transcription error. Table 3 publishes only the arm **marginals** -- the
HBeAg-positive percentage and the mean baseline HBsAg -- while the paper
simulated their joint distribution, in which HBeAg-positive patients carry
higher baseline HBsAg. Recombining from marginals assumes independence, which
biases the weighted average upward. The signature confirms the explanation: the
bias is positive on every mixed row (max
`r sprintf("%.2f", max(mixed_ok$eot_diff))` pp) and vanishes on both 0%-HBeAg
rows, where there is no mixture to approximate.

# Check 5 -- between-trial variance reduction

Results Section 3.3 reports the variance reduction achieved by adding the
treatment and covariate effects to the base models.

```{r check5-variance-reduction}
check5 <- tibble::tibble(
  model      = c("EOT", "24-week post-treatment"),
  base_var   = c(1.112, 2.544),          # Results Section 3.3
  final_var  = c(omega_eot, omega_24wk), # encoded in the model files
  results_pct  = c(56.5, 77.5),          # Results Section 3.3
  abstract_pct = c(58.1, 77.6)           # Abstract and Discussion
) |>
  mutate(computed_pct = 100 * (base_var - final_var) / base_var)

check5 |>
  dplyr::rename(
    "Model"                        = model,
    "Base-model variance"          = base_var,
    "Final-model variance"         = final_var,
    "Computed reduction (%)"       = computed_pct,
    "Results Section 3.3 (%)"      = results_pct,
    "Abstract / Discussion (%)"    = abstract_pct
  ) |>
  knitr::kable(digits = 2, caption = "Between-trial variance reduction. The Results figures are arithmetically correct; the Abstract and Discussion disagree (see Errata).")

# The encoded final variances must reproduce the Results Section 3.3 figures.
stopifnot(abs(check5$computed_pct - check5$results_pct) < 0.1)
# ...and must NOT reproduce the Abstract's EOT figure, which is the erratum.
stopifnot(abs(check5$computed_pct[1] - check5$abstract_pct[1]) > 1)
```

# Check 6 -- Table 3's flagged row is self-falsified by the paper's own coefficients

Each Table 3 row can be back-solved: invert the 24-week model at the row's
published rate to recover the baseline HBsAg the paper must have simulated, and
compare it to the baseline the row prints.

```{r check6-backsolve}
logit <- function(p) log(p / (1 - p))

backsolve_bl <- function(rate_pct, conmed_nuc) {
  intercept <- if (conmed_nuc == 1) getpar(ui_24wk, "logitp_comb") else getpar(ui_24wk, "logitp_mono")
  # duration is 48 weeks throughout Table 3, so the duration term is zero
  3 + (logit(rate_pct / 100) - intercept) / getpar(ui_24wk, "e_hbsag_bl_log10_logitp")
}

check6 <- check4 |>
  rowwise() |>
  mutate(implied_bl = backsolve_bl(wk24_pub, regimen == "Peg-IFNa + NA")) |>
  ungroup() |>
  mutate(bl_discrepancy = implied_bl - mean_bl)

check6 |>
  transmute(regimen, criteria, mean_bl, implied_bl, bl_discrepancy) |>
  dplyr::rename(
    "Treatment group"                    = regimen,
    "HBsAg criteria (log10)"             = criteria,
    "Printed mean BL HBsAg (log10)"      = mean_bl,
    "Back-solved BL HBsAg (log10)"       = implied_bl,
    "Discrepancy (log10)"                = bl_discrepancy
  ) |>
  knitr::kable(digits = 3, caption = "Back-solving each Table 3 row's 24-week rate through the paper's own Table 2 coefficients.")

consistent <- check6 |> filter(row != flagged$row)
stopifnot(abs(consistent$bl_discrepancy) < 0.025)

# The flagged row is an order of magnitude further out than any other row.
flagged_disc <- check6$bl_discrepancy[check6$row == flagged$row]
stopifnot(abs(flagged_disc) > 0.15,
          abs(flagged_disc) > 5 * max(abs(consistent$bl_discrepancy)))
```

Seven of the eight rows back-solve to within
`r sprintf("%.3f", max(abs(consistent$bl_discrepancy)))` log10 IU/mL of their
printed baseline. The Peg-IFN-alpha + NA "<= 5 log10" row does not: its printed
3.3 implies `r sprintf("%.2f", check6$implied_bl[check6$row == flagged$row])`
from its own published 6.8% rate. Either the baseline or the rate in that one
row is mis-transcribed, and the paper gives no way to tell which.

# Covariate response surface

The paper's central claim is that lower baseline HBsAg and longer treatment
duration both raise the probability of HBsAg loss.

```{r figure-surface, fig.cap = "Study-arm HBsAg-loss probability against baseline HBsAg for three Peg-IFN-alpha durations, both endpoints, both regimens (typical study-arm, HBeAg-negative)."}
grid_arms <- tidyr::crossing(
  HBSAG_BL_LOG10 = seq(1.5, 4.5, by = 0.1),
  T_PEGIFN       = c(24, 48, 96),
  CONMED_NUC     = c(0, 1)
) |>
  mutate(HBEAG_POS = 0)

surface <- bind_rows(
  solve_arms(mod_eot_typ, grid_arms,
             keep = c("HBSAG_BL_LOG10", "T_PEGIFN", "CONMED_NUC")) |>
    mutate(endpoint = "End of treatment"),
  solve_arms(mod_24wk_typ, grid_arms,
             keep = c("HBSAG_BL_LOG10", "T_PEGIFN", "CONMED_NUC")) |>
    mutate(endpoint = "24 weeks post-treatment")
) |>
  mutate(
    regimen  = ifelse(CONMED_NUC == 1, "Peg-IFNa + NA", "Peg-IFNa monotherapy"),
    duration = factor(T_PEGIFN, levels = c(24, 48, 96),
                      labels = c("24 weeks", "48 weeks", "96 weeks")),
    endpoint = factor(endpoint,
                      levels = c("End of treatment", "24 weeks post-treatment"))
  )

ggplot(surface, aes(x = HBSAG_BL_LOG10, y = p_hbsag_loss, colour = duration)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 3, linetype = "dashed", alpha = 0.5) +
  facet_grid(regimen ~ endpoint) +
  scale_y_continuous(labels = scales::percent_format(1)) +
  labs(
    x = "Baseline HBsAg (log10 IU/mL)",
    y = "Study-arm HBsAg-loss probability",
    colour = "Peg-IFNa duration",
    caption = "Dashed line: the 3 log10 IU/mL reference. Typical study-arm (eta_study = 0), HBeAg-negative."
  ) +
  theme_bw() +
  theme(legend.position = "top")
```

The duration effect is visibly larger at 24 weeks post-treatment than at end of
treatment, which is the Table 2 result that the per-week odds ratio rises from
`r sprintf("%.3f", check2$model_or[1])` to
`r sprintf("%.3f", check2$model_or[4])`.

# Validation strategy: why there is no PKNCA section

The standard popPK validation for this library is a PKNCA non-compartmental
analysis against the paper's reported Cmax / AUC / half-life. That is not
applicable here and is deliberately omitted:

* There is no pharmacokinetic component. Neither model contains a
  concentration, a dose amount, a volume or a clearance -- the Peg-IFN-alpha
  regimen enters only as a binary combination indicator and a planned-duration
  covariate, because non-standard Peg-IFN-alpha dosing was an exclusion
  criterion.
* The model output `Cc` is a dimensionless **proportion** in [0, 1], not a
  concentration. It is named `Cc` only to satisfy the library's endpoint
  convention.
* There is no time axis to integrate over: both models are cross-sectional.

The validation above substitutes the checks that can actually catch a
translation error in this model class: exact reproduction of every published
parameter back-transform (Check 1), odds ratios recovered from the solved model
rather than from the inputs (Check 2), reconstruction of the published
prediction intervals that discriminates a variance from a standard deviation
(Check 3), the paper's own simulation answer key (Check 4), the variance-
reduction arithmetic (Check 5), and an internal-consistency back-solve of the
answer key against the parameter table (Check 6).

# Assumptions and deviations

* **Two model files, one vignette.** The paper fits its endpoints independently
  (Section 2.3) on different analysis sets with different retained covariates,
  so the author structure is replicated as two `.R` files sharing this vignette.
* **`addSd <- fixed(0.001)` is not a paper parameter.** Section 2.3.1 fixes the
  residual variance to 1 with a binomial variance function, so there is no
  estimated residual-error term. The placeholder exists only so the model
  declares an nlmixr2 endpoint; it is negligible and the paper's own Table 3
  simulations exclude any residual term (Section 2.5 resamples fixed effects and
  the study eta only).
* **`eta_study` is a paper-specific eta name.** It is a between-study-strata-arm
  effect, not a between-subject effect on a structural PK parameter, so it is
  declared in `paper_specific_etas` rather than renamed to an `etal<param>`
  form.
* **Simulation scope is the study arm.** Every output here is an aggregate
  study-arm proportion. These models must not be used for individual-patient
  simulation.
* **HBeAg mixture handling.** For an arm containing both HBeAg strata, simulate
  the strata separately and take the HBeAg-weighted average, as Check 4 does.
  Evaluating the model at a fractional `HBEAG_POS` is wrong because the model is
  logit-linear in the indicator.
* **Screened-but-not-retained covariates carry no estimates.** Age, gender,
  race, study design, and (24-week model only) continuation of NA after
  Peg-IFN-alpha cessation were tested and not retained; the paper reports no
  point estimate for any of them. They are documented in
  `covariatesDataExcluded` in each model file rather than encoded.
* **Race orientation.** The paper's race covariate is the proportion
  *non-Asian*; the register's canonical `RACE_ASIAN_PCT` is the complement. This
  matters only if someone revives the screened covariate, since it is not in
  either final model.
* **Base models are not extracted.** The base models (single overall treatment
  effect, EOT 6.7% [95% PI 0.9-36.7] and 24-week 6.7% [95% PI 0.3-63.7]) are
  reported only through those rates and their between-trial variances; their
  intercepts are not tabulated, so they are not reproducible and only the final
  covariate models are extracted.

# Errata

Discrepancies found in the source while verifying. None changes an encoded
parameter value; each is recorded so a reader comparing the model file against
the paper is not surprised.

1. **HBeAg p-value.** Table 2 prints `p = 0.008` for the EOT HBeAg-status term;
   the Abstract and Results Section 3.3 both say `p = 0.007`. Estimates are
   unaffected.
2. **Between-trial variance reduction.** Results Section 3.3 gives 56.5% (EOT)
   and 77.5% (24-week), which Check 5 confirms is the arithmetically correct
   pair. The Abstract and Discussion instead give 58.1% and 77.6%. The Results
   figures are the right ones; note this is the reverse of the usual pattern, in
   which the Results narrative is the one that drifts.
3. **Table 3, Peg-IFN-alpha + NA "<= 5 log10" row is internally inconsistent.**
   Check 6 back-solves its published 24-week rate of 6.8% through the paper's
   own Table 2 coefficients and recovers a baseline of
   `r sprintf("%.2f", check6$implied_bl[check6$row == flagged$row])` log10 IU/mL
   against a printed 3.3. Every other row back-solves to within
   `r sprintf("%.3f", max(abs(consistent$bl_discrepancy)))`. Either the baseline
   or the rate in that one row is mis-transcribed and the paper gives no way to
   tell which, so the row is excluded from the Check 4 and Check 6 assertions
   and is flagged in both tables.
4. **Table 2 cosmetic slips.** The 24-week "Drug effect (Peg-IFNa + NA)"
   Interpretation cell has a doubled percent sign ("14.1% %") and names the
   regimen "48 weeks Peg-IFNa" without the "+ NA".
5. **Section 3.5 rates are not reproducible from Table 2.** The in-silico-trial
   benchmarks (15.1% monotherapy, 19.1% combination, for virtual cohorts
   calibrated at baseline HBsAg 2485 and 2081 IU/mL) cannot be recovered from
   the Table 2 coefficients at any plausible duration or HBeAg mix; at
   2485 IU/mL = 3.40 log10 and 48 weeks the models give roughly 4.8% (EOT) and
   6.3% (24-week). Those figures come from the companion Cortes-Rios 2025
   in-silico-trial calibration rather than from this paper's own models, so they
   are not used as a validation target. Note also that Section 3.5 is where the
   paper quotes baselines in **raw IU/mL** while the models take **log10
   IU/mL** -- the reason the canonical covariate is named `HBSAG_BL_LOG10`.
