---
title: "Rosuvastatin (Hanke 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rosuvastatin (Hanke 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Hanke 2021 is mainly a whole-body PBPK paper. It builds a PK-Sim model of
rosuvastatin with OATP2B1, OATP1B1/1B3, OAT3, P-gp and BCRP transport plus
CYP2C9 metabolism, and uses it to predict the rosuvastatin drug-drug
interactions (DDIs) with rifampicin, gemfibrozil and probenecid. Rosuvastatin
plasma profiles rise slowly and peak late (median tmax 5.0 h). To describe this,
the authors ran a **NONMEM population-PK analysis of rosuvastatin**, reported in
Electronic Supplementary Material (ESM) section 2 with the final control stream
in section 2.4.2. The PBPK oral dosing protocols took their split-dose input
(63.4% of the dose at 0 h, 36.6% at 2.3 h) from this analysis.

That population-PK analysis is what this package provides. The PK-Sim
whole-body PBPK layer is not reproduced. Its organ volumes, blood flows,
partition coefficients and transporter expression are PK-Sim database outputs
that appear nowhere in the paper.

```{r model}
mod <- readModelDb("Hanke_2021_rosuvastatin")
```

* Citation: `r rxode2::rxode2(mod)$reference`
* Article: <https://doi.org/10.1007/s11095-021-03109-6>
* ESM 1 (model documentation, Tables S2.3.1-S2.4.1, the NONMEM code, and the
  observed study AUClast / Cmax in Table S3.5.2): distributed with the article
  and available through Europe PMC (`PMC8602162`, supplementary file
  `11095_2021_3109_MOESM1_ESM.pdf`).

`r rxode2::rxode2(mod)$description`

# Population

ESM Table S2.3.1 lists the three studies used to build the population-PK
model:

| Study | Design | Profiles | Age (years) | Weight (kg) |
|---|---|---|---|---|
| Martin et al. 2003c | 8 mg IV, 4-h infusion (n = 10) | 1 (digitised mean) | 36 (21-51) | 78 (68-85) |
| Stopfer et al. 2016 | 10 mg oral tablet, fasted | 19 | 37 (23-49) | 85 (68-99) |
| Stopfer et al. 2018b | 10 mg oral tablet, fasted | 25 | 35 (20-55) | 84 (67-105) |

ESM Table S3.2.1 lists all three as 100% male European cohorts of healthy
volunteers. The model has no covariates.

# Source trace

| Element | Value | Source |
|---|---|---|
| Two-compartment disposition, first-order elimination from central | -- | ESM section 2.4.1; `$DES` in section 2.4.2 |
| Oral dose split into two portions, shared `KA1`, lag on the second only | -- | ESM section 2.4.1, Figure S2.4.1; `$PK`/`$DES` |
| `lka` | 0.464 1/h | Table S2.4.1 `Ka` (without DDI) |
| `lfdepot` (FTOT) | 0.0793 | Table S2.4.1 `Ftot` = 7.93% |
| `logitfrac` (VF2) | logit(0.366) | Table S2.4.1 `VF2`; `$PK` `PHI_2 = LOG(VF2/(1-VF2))` |
| `ltlag2` (ALAG2) | 2.30 h | Table S2.4.1 `ALAG2` |
| `lcl` | 19.1 L/h | Table S2.4.1 `CL` |
| `lvc` | 79.4 L | Table S2.4.1 `V3` |
| `lq` | 12.0 L/h | Table S2.4.1 `Q` |
| `lvp` | 199 L | Table S2.4.1 `V4` |
| `etalogitfrac` | log(1 + 0.778^2) | Table S2.4.1 `IIV VF2` = 77.8 %CV (logit scale, `$PK` `VF2_2`) |
| `etalcl` | log(1 + 0.261^2) | Table S2.4.1 `IIV CL` = 26.1 %CV |
| `etalfdepot` | log(1 + 0.801^2) | Table S2.4.1 `IIV Ftot` = 80.1 %CV; on `F1` only (`$PK` `F1 = VF1*FTOT*EXP(ETA(3))`) |
| `propSd` | 0.222 | Table S2.4.1 `Prop RE` = 22.2% |
| `addSd` | 0.0077460 ng/mL, fixed | `$SIGMA` `0.00006 FIX`, square-rooted; Table S2.4.1 `Add RE` 0.00775 ng/ml (fixed) |
| `Cc = 1000 * central / vc` | ng/mL | `$PK` `S3 = V3/1000` |
| `Cc ~ add + prop` | -- | `$ERROR` `Y = IPRED + W*EPS(1) + EPS(2)`, `W = F` |

# Dosing convention

The NONMEM model splits an oral dose by bioavailability fractions:
`F1 = (1 - VF2) * FTOT * exp(ETA(3))` for `DEPOT1` and `F2 = VF2 * FTOT` for
`DEPOT2`, with `ALAG2` on `DEPOT2` only. An oral dose is therefore given as
**two dose records of the full dose at the same time**, one to `depot` and one
to `depot2`. An intravenous dose is a single record to `central`.

```{r events}
obs_times <- sort(unique(c(seq(0, 12, by = 0.1), seq(12.5, 24, by = 0.5), seq(25, 72, by = 1))))

oral_events <- function(id, dose, treatment) {
  dplyr::bind_rows(
    data.frame(id = id, time = 0, amt = dose, evid = 1, cmt = c("depot", "depot2"), rate = 0),
    data.frame(id = id, time = obs_times, amt = 0, evid = 0, cmt = "central", rate = 0)
  ) |>
    dplyr::mutate(treatment = treatment)
}

# 8 mg over 4 h, as in Martin et al. 2003c.
iv_events <- function(id, dose, duration, treatment) {
  dplyr::bind_rows(
    data.frame(id = id, time = 0, amt = dose, evid = 1, cmt = "central", rate = dose / duration),
    data.frame(id = id, time = obs_times, amt = 0, evid = 0, cmt = "central", rate = 0)
  ) |>
    dplyr::mutate(treatment = treatment)
}
```

# Typical-value simulation

```{r typical}
mod_typ <- rxode2::zeroRe(mod)

ev_typ <- dplyr::bind_rows(
  oral_events(1, 10, "10 mg oral"),
  iv_events(2, 8, 4, "8 mg IV (4-h infusion)")
)

# Tight tolerances: these solves feed the closed-form checks below.
sim_typ <- rxode2::rxSolve(
  mod_typ, ev_typ,
  keep = "treatment", returnType = "data.frame",
  rtol = 1e-10, atol = 1e-12
)
```

```{r typical-plot, fig.cap = "Typical-value rosuvastatin profiles. The shoulder in the oral profile at 2.3 h is the delayed second dose portion (compare Figure 1 of Hanke 2021)."}
ggplot(sim_typ, aes(time, Cc, colour = treatment)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Rosuvastatin (ng/mL)", colour = NULL)
```

# Stochastic simulation

A cohort of 200 virtual subjects receives a single 10 mg fasted oral dose, the
design of both oral training studies.

```{r vpc}
n_sub <- 200
ev_vpc <- dplyr::bind_rows(lapply(seq_len(n_sub), function(i) {
  oral_events(i, 10, "10 mg oral")
}))
rxode2::rxSetSeed(20211018)
sim_vpc <- rxode2::rxSolve(mod, ev_vpc, keep = "treatment", returnType = "data.frame")

vpc_band <- sim_vpc |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    q05 = quantile(Cc, 0.05),
    q50 = quantile(Cc, 0.50),
    q95 = quantile(Cc, 0.95),
    .groups = "drop"
  )
```

```{r vpc-plot, fig.cap = "Median and 90% interval of 200 simulated subjects after 10 mg rosuvastatin orally (IPRED, without residual error). Dashed lines: observed Cmax of Stopfer et al. 2016 (4.48 ng/mL) and 2018b (4.11 ng/mL), ESM Table S3.5.2. Compare with the individual fits in ESM Figures S2.4.3-S2.4.6."}
ggplot(vpc_band, aes(time)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), fill = "steelblue", alpha = 0.3) +
  geom_line(aes(y = q50), colour = "steelblue") +
  geom_hline(yintercept = c(4.48, 4.11), linetype = "dashed", colour = "grey40") +
  coord_cartesian(xlim = c(0, 48)) +
  labs(x = "Time (h)", y = "Rosuvastatin (ng/mL)")
```

# PKNCA validation

```{r pknca}
nca_for <- function(sim, events, conc_col) {
  conc <- sim |>
    dplyr::mutate(conc = .data[[conc_col]]) |>
    dplyr::filter(!is.na(conc)) |>
    dplyr::select(id, time, conc, treatment)
  # One dose row per id: for oral dosing the depot and depot2 records carry
  # the same full dose.
  dose <- events |>
    dplyr::filter(evid == 1) |>
    dplyr::distinct(id, time, amt, treatment)
  PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(conc, conc ~ time | treatment + id),
    PKNCA::PKNCAdose(dose, amt ~ time | treatment + id),
    intervals = data.frame(
      start = 0, end = Inf,
      cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE
    )
  ))
}

nca_typ <- as.data.frame(nca_for(sim_typ, ev_typ, "Cc")) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "aucinf.obs", "half.life")) |>
  dplyr::select(treatment, PPTESTCD, PPORRES)

nca_vpc <- as.data.frame(nca_for(sim_vpc, ev_vpc, "Cc")) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast")) |>
  dplyr::select(treatment, id, PPTESTCD, PPORRES)
```

## Structural checks

The typical-value AUC0-inf has a closed form, `Dose / CL` intravenously and
`FTOT * Dose / CL` orally, whatever the split fraction and lag. The terminal
half-life is `ln 2 / beta`, where `beta` is the smaller eigenvalue of the
two-compartment system. Both sides use the same parameters, so the only
difference is numerical error in the solve and in the NCA extrapolation, and a
tight bound is appropriate.

```{r structural}
p <- list(cl = 19.1, vc = 79.4, q = 12.0, vp = 199, f = 0.0793)
k10 <- p$cl / p$vc
k12 <- p$q / p$vc
k21 <- p$q / p$vp
beta <- ((k10 + k12 + k21) - sqrt((k10 + k12 + k21)^2 - 4 * k10 * k21)) / 2

expected <- tibble::tibble(
  treatment = c("10 mg oral", "8 mg IV (4-h infusion)"),
  aucinf_expected = c(p$f * 10, 8) / p$cl * 1000,
  half_life_expected = log(2) / beta
)

check_typ <- nca_typ |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::left_join(expected, by = "treatment")

stopifnot(
  all(abs(check_typ$aucinf.obs / check_typ$aucinf_expected - 1) < 0.01),
  all(abs(check_typ$half.life / check_typ$half_life_expected - 1) < 0.02)
)

# The paper's split: 63.4% of the dose without lag, 36.6% after 2.3 h.
stopifnot(
  abs(plogis(log(0.366 / (1 - 0.366))) - 0.366) < 1e-12,
  abs(1 - 0.366 - 0.634) < 1e-12
)

knitr::kable(
  check_typ |>
    dplyr::select(treatment, cmax, tmax, aucinf.obs, aucinf_expected, half.life, half_life_expected) |>
    dplyr::rename(
      "Treatment" = treatment,
      "Cmax (ng/mL)" = cmax,
      "Tmax (h)" = tmax,
      "AUC0-inf (ng*h/mL)" = aucinf.obs,
      "Closed-form AUC0-inf" = aucinf_expected,
      "t1/2 (h)" = half.life,
      "Closed-form t1/2" = half_life_expected
    ),
  digits = 2,
  caption = "Typical-value NCA against the closed forms."
)
```

## Comparison against published NCA

ESM Table S3.5.2 gives the observed AUClast and Cmax of every study in the
paper; its "Pred" columns are PBPK predictions and are not used here. The two
oral training studies of the population-PK model are compared with the median
of the simulated 10 mg cohort. The simulated AUClast runs to 72 h. The ESM does
not give the last sampling time, but the typical-value AUC grows by only about
2.5% between 72 and 96 h, so the choice hardly matters.

```{r nca-compare}
published <- tibble::tribble(
  ~study,                 ~cmax, ~auclast,
  "Stopfer et al. 2016",  4.48,  41.00,
  "Stopfer et al. 2018b", 4.11,  42.22
)

sim_nca <- nca_vpc |>
  dplyr::select(-id)

cmp <- dplyr::bind_rows(lapply(seq_len(nrow(published)), function(i) {
  nlmixr2lib::ncaComparisonTable(
    simulated = sim_nca |> dplyr::mutate(treatment = published$study[i]),
    reference = published[i, ] |> dplyr::rename(treatment = study),
    by = "treatment",
    units = c(cmax = "ng/mL", tmax = "h", auclast = "ng*h/mL"),
    tolerance_pct = 20
  )
}))

knitr::kable(
  cmp,
  caption = "Median of 200 simulated subjects (10 mg oral) vs. the observed values of the two oral training studies in Hanke 2021 ESM Table S3.5.2. * differs from the reference by more than 20%.",
  align = c("l", "l", "r", "r", "r")
)
```

```{r nca-gate}
vpc_wide <- nca_vpc |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

r_auc <- median(vpc_wide$auclast) / published$auclast
r_cmax <- median(vpc_wide$cmax) / published$cmax

stopifnot(
  # AUC depends only on FTOT / CL for the oral route: a mis-transcribed
  # clearance, bioavailability or unit moves it by tens of percent.
  all(abs(r_auc - 1) < 0.2),
  # Cmax depends on ka, the split, the lag and the volumes; the observed
  # study summaries are not stated to be medians, so the tolerance is wider.
  all(abs(r_cmax - 1) < 0.3),
  # The median tmax reproduces the late peak described in the paper
  # (tmax = 5.0 h in the Results; typical-value 3.8 h here).
  median(vpc_wide$tmax) > 2.5,
  median(vpc_wide$tmax) < 6
)
```

The simulated median **AUClast** is within a few percent of both observed
values. The simulated median **Cmax** runs below the observed values by about
8-15%. The observed values in Table S3.5.2 are study summaries (their
statistic is not stated), and the 80% CV on the first-portion bioavailability
makes a mean Cmax sit above the median.

## Intravenous arm: a documented discrepancy

The same table gives the observed AUClast of the intravenous training study,
Martin et al. 2003c (8 mg over 4 h): 163.14 ng*h/mL. The model's typical-value
AUC0-inf for that dose is `8 / 19.1 * 1000` = 419 ng*h/mL, about 2.5 times
higher. The observed data alone imply a bioavailability of about 0.20
(dose-normalised oral over intravenous AUClast), against the fitted FTOT of
0.0793. The oral data pin only `FTOT / CL`. The printed CL and FTOT are both
about 2.5-fold lower than the data-only values, and their ratio still matches,
which is why the oral comparisons above agree while the intravenous one does
not. The ESM does not show the population-PK fit to the intravenous profile
(Figures S2.4.3-S2.4.8 show only the oral and DDI subjects), so the cause
cannot be identified from the paper. The model is provided as published.

```{r iv-check}
iv_auc_obs <- 163.14
oral_auc_obs <- mean(published$auclast)
iv_sim <- check_typ$aucinf.obs[check_typ$treatment == "8 mg IV (4-h infusion)"]
f_data <- (oral_auc_obs / 10) / (iv_auc_obs / 8)

knitr::kable(
  tibble::tibble(
    Quantity = c(
      "Typical IV AUC0-inf, model (ng*h/mL)",
      "Observed IV AUClast, Martin et al. 2003c (ng*h/mL)",
      "Ratio model / observed",
      "Data-only bioavailability (oral / IV, dose-normalised)",
      "Fitted FTOT"
    ),
    Value = c(iv_sim, iv_auc_obs, iv_sim / iv_auc_obs, f_data, 0.0793)
  ),
  digits = 3,
  caption = "The intravenous arm is not reproduced by the published CL and FTOT."
)

# This locks in the documented discrepancy (deterministic typical-value
# solve). If the parameter values change, update this section.
stopifnot(iv_sim / iv_auc_obs > 2, iv_sim / iv_auc_obs < 3)
```

# The DDI extension of the population-PK model

ESM section 2.4.1 also refits the model with the rifampicin, probenecid and
gemfibrozil DDI study arms added. Covariate factors on bioavailability and
clearance describe each interaction. During rifampicin and probenecid
co-administration the dose is not split and has no lag time. The re-estimated
typical values ("with DDI" column of Table S2.4.1) are shown below for
reference. The DDI covariate factors themselves are not reported anywhere in
the paper or ESM, and the control stream in section 2.4.2 is the model without
DDI, so the DDI model is not provided.

| Parameter | Without DDI (this model) | With DDI |
|---|---|---|
| Ka (1/h) | 0.464 | 0.397 |
| Ftot (%) | 7.93 | 8.40 |
| VF2 | 0.366 | 0.338 |
| ALAG2 (h) | 2.30 | 2.26 |
| CL (L/h) | 19.1 | 18.9 |
| V3 (L) | 79.4 | 83.2 |
| Q (L/h) | 12.0 | 12.5 |
| V4 (L) | 199 | 211 |
| IIV Ftot (%CV) | 80.1 | 69.8 |
| IIV VF2 (%CV) | 77.8 | 64.8 |
| IIV CL (%CV) | 26.1 | 27.1 |
| Prop RE (%) | 22.2 | 26.9 |
| Add RE (ng/mL) | 0.00775 (fixed) | n.a. |

# Assumptions and deviations

* **Only the monotherapy population-PK model is provided.** The PK-Sim
  whole-body PBPK model (including all DDI predictions) is not reproduced. Its
  partition coefficients, organ physiology and transporter expression are
  platform database outputs that the paper does not tabulate. The DDI
  extension of the population-PK model is not provided because its covariate
  factors are not reported (see above).
* **IIV "Ftot" acts on the first dose portion only.** Table S2.4.1 and the text
  describe IIV on the total bioavailability. The control stream applies
  `EXP(ETA(3))` to `F1` only (`F2 = VF2_2 * FTOT` has no `ETA(3)`). The model
  follows the control stream.
* **%CV to variance.** Table S2.4.1 reports each IIV as a %CV. The variances
  here use `omega^2 = log(1 + CV^2)`, the log-normal relation the ESM itself
  uses elsewhere (Table S7.0.1 footnote: "35 % CV was assumed (= 1.40
  GeoSD)"). The same conversion is applied to the logit-scale VF2 eta. If the
  authors instead reported `100 * sqrt(omega^2)` for that row, the variance
  would be 0.605 rather than 0.473. The `$OMEGA` values printed in the control
  stream are initial estimates and were not used.
* **Additive residual error.** The control stream fixes the additive `$SIGMA`
  at 0.00006, i.e. an SD of 0.0077 ng/mL, matching Table S2.4.1.
* **Concentration unit.** The control stream scales the central amount by
  `S3 = V3/1000` with volumes in litres. With doses in mg this gives ng/mL,
  the unit of the additive error and of the observed values in Table S3.5.2.
* **Intravenous arm.** The published CL and FTOT do not reproduce the observed
  intravenous AUClast (see above). Oral predictions, which depend on
  `FTOT / CL`, agree with the observed data. No parameter was adjusted.
* **Dose records.** An oral dose must be given as two records of the full dose,
  to `depot` and `depot2`. A single record to `depot` alone would deliver only
  `(1 - VF2) * FTOT` of the dose.
