---
title: "Bortezomib (Hanley 2017)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bortezomib (Hanley 2017)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r load-model}
ui <- rxode2::rxode(readModelDb("Hanley_2017_bortezomib"))
```

# Model and source

* Citation: `r ui$reference`
* Description: `r ui$description`
* Article: <https://doi.org/10.1002/jcph.906>
* Open-access full text: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5561493/>

# Population

The model was fit to 571 bortezomib plasma concentrations from 104 pediatric
patients aged 2 to 16 years enrolled in two Children's Oncology Group trials
(Table 1 of Hanley 2017): **AALL07P1** (NCT00873093, phase 2, relapsed acute
lymphoblastic leukemia, N = 51) and **AAML1031** (NCT01371981, phase 3, de novo
acute myelogenous leukemia, N = 53). Sixty percent of patients were 2-11 years
old and 40% were 12-16 years old; the mean age was 10 years. Body surface area
spanned 0.60-2.53 m^2 (mean 1.30 m^2) and body weight 13.9-139.7 kg (mean
45.2 kg). Fifty-eight percent were male, and 65% were White, 18% Black and 17%
Other. All patients received bortezomib 1.3 mg/m^2 as an intravenous bolus
twice weekly (days 1, 4 and 8 of the relevant blocks or courses, plus day 11 of
block 1 of AALL07P1) on top of multiagent chemotherapy backbones.

Pharmacokinetic sampling was confined to the **day 8** dose over 0-72 h
postdose. AALL07P1 sampled predose, 5-15 min, 30-60 min, 4-8 h and ~72 h;
AAML1031 additionally sampled 18-30 h. The median was 6 concentrations per
patient (range 1-6) and no concentration fell below the 0.1 ng/mL limit of
quantification.

Two scope limits stated by the authors carry into any use of this model. First,
the fit used **repeat-dose** data only, and bortezomib clearance is known to
fall between the first dose of cycle 1 and steady state, so these parameters
must not be used to describe first-dose exposure. Second, patients under
2 years of age were not studied.

The same information is available programmatically via the model's `population`
metadata:

```{r population}
str(ui$population, max.level = 1)
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Hanley_2017_bortezomib.R` carries an in-file
comment naming its source location. They are collected here for review.

| Equation / parameter | Value as encoded | Source location |
|---|---|---|
| Three-compartment linear disposition, IV bolus into `central` | n/a | Results: "a 3-compartment model was selected as the base model"; Methods "Population PK Modeling" |
| `lcl` (CL) | `log(9.59)` L/h | Table 2, row "CL, L/h" = 9.59 (95%CI 8.79-10.37) |
| `e_bsa_cl` | 0.97 | Table 2, row "BSA effect on CL" = 0.97 (95%CI 0.72-1.25) |
| `lvc` (V1) | `log(10.0)` L | Table 2, row "V1, L" = 10.0 (95%CI 6.09-13.4) |
| `lq` (Q2) | `log(25.8)` L/h | Table 2, row "Q2, L/h" = 25.8 (95%CI 18.9-31.9) |
| `lvp` (V2) | `log(32.5)` L | Table 2, row "V2, L" = 32.5 (95%CI 23.1-43.1) |
| `lq2` (Q3) | `log(26.6)` L/h | Table 2, row "Q3, L/h" = 26.6 (95%CI 21.3-30.7) |
| `e_bsa_q2` | 0.75 | Table 2, row "BSA effect on Q3" = 0.75 (95%CI 0.43-0.99) |
| `lvp2` (V3) | `log(975)` L | Table 2, row "V3, L" = 975 (95%CI 792-1190) |
| BSA normalisation constant | 1.30 m^2 | Table 1, total-cohort "Mean BSA, m2" = 1.30; see "Assumptions and deviations" for why this, and not the Methods' "1.8 m2", is the correct constant |
| `etalcl` | `log(1 + 0.297^2)` | Table 2 CL IIV = 29.7 %CV; Results "CL 29.7%" |
| `etalvc` | `log(1 + 0.346^2)` | Table 2 V1 IIV = 34.6 %CV; Results "V1 34.6%" |
| `etalq2` | `log(1 + 0.298^2)` | Table 2 Q3 IIV = 29.8 %CV; Results "Q3 29.8%" |
| Exponential (log-normal) IIV | n/a | Methods: "estimate population PK parameters and associated interindividual variability (IIV) using an exponential error model" |
| `expSdAall07p1` | `sqrt(log(1 + 0.468^2))` | Table 2, row "Residual error for study AALL07P1,%CV" = 46.8% |
| `expSdAaml1031` | `sqrt(log(1 + 0.219^2))` | Table 2, row "Residual error for study AAML1031,%CV" = 21.9% |
| Log-additive (`lnorm`) residual, split by study | n/a | Methods: "the log-transform both sides approach for residual variability"; Results: "The residual error model was a homoscedastic model on the log scale, proportional after back transformation" and "separate residual variability parameters were estimated for each study" |
| Covariates screened but not retained (age, weight, race, disease type, AML risk group, ALL treatment plan stratum) | n/a, carried in `covariatesDataExcluded` | Methods "Population PK Modeling"; Results: "No additional statistically significant covariates were identified" |

# Virtual cohort

Original patient-level data are not publicly available. The cohort below
reproduces the per-study body surface area distributions of Table 1: mean
1.20 m^2 (range 0.60-2.21) for AALL07P1 and mean 1.40 m^2 (range 0.61-2.53) for
AAML1031, with N = 51 and N = 53 respectively, matching the published study
sizes. BSA is drawn log-normally with a 35% coefficient of variation, which is
what the published mean and range imply, and clamped to the reported range.

The register entry for `BSA` warns that a BSA-power exponent is estimated
against whatever BSA scale the source paper used, so the cohort must sample BSA
from the reported distribution rather than recomputing it from height and
weight. Hanley 2017 does not state which BSA formula was used, so that is the
only defensible route here in any case.

```{r cohort}
# set.seed() seeds R's RNG, which is what draws BSA below, so the covariate
# cohort IS reproducible across machines. It does NOT seed rxode2's simulation
# RNG (the etas), whose streams are partitioned per solver thread -- so every
# assertion on an eta-derived quantity downstream is written to hold for any
# cohort the model can produce.
set.seed(20170908)

draw_bsa <- function(n, mean_bsa, lo, hi, cv = 0.35) {
  sdlog <- sqrt(log(1 + cv^2))
  meanlog <- log(mean_bsa) - sdlog^2 / 2
  x <- stats::rlnorm(n, meanlog, sdlog)
  # Rescale to the published mean before clamping. Without this the sampling
  # error of a 51-subject mean at 35% CV is ~5%, which propagates into every
  # clearance-derived comparison below (one draw came in 6.9% low and pushed
  # that arm's terminal half-life to +12.6%). Calibrating the draw removes a
  # noise source that has nothing to do with the transcription being tested.
  pmin(pmax(x * mean_bsa / mean(x), lo), hi)
}

make_cohort <- function(n, bsa, study_flag, treatment, dose_days, id_offset = 0L) {
  subj <- tibble(
    id = id_offset + seq_len(n),
    BSA = bsa,
    STUDY_AALL07P1 = study_flag,
    treatment = treatment
  )
  # 1.3 mg/m2 IV bolus into the central compartment on the listed study days.
  doses <- subj |>
    tidyr::crossing(day = dose_days) |>
    mutate(
      time = (day - 1) * 24,
      amt = 1.3 * BSA,
      evid = 1L,
      cmt = "central"
    ) |>
    select(-day)
  # Observations over the day-8 dose, 0-72 h postdose (the paper's PK window).
  # Nominal sample times plus a dense grid for the figure. Capped just short of
  # 72 h so the last record cannot collide with the AALL07P1 day-11 dose.
  obs_times <- sort(unique(c(
    168, 168 + c(1 / 6, 0.75, 6, 24, 71.9),
    seq(168, 239.9, length.out = 150)
  )))
  obs <- subj |>
    tidyr::crossing(time = obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}

# Published per-study demographics (Hanley 2017 Table 1). Everything
# downstream is derived from this one table and joined BY NAME, so no check
# can silently pair a study with the other study's published value.
study_tab <- tibble::tribble(
  ~treatment,                ~study_flag, ~n,  ~published_mean, ~lo,  ~hi,
  "AALL07P1 (relapsed ALL)", 1,           51L, 1.20,            0.60, 2.21,
  "AAML1031 (de novo AML)",  0,           53L, 1.40,            0.61, 2.53
)

# Bortezomib dosing days, keyed by treatment label: days 1, 4, 8 and 11 of
# block 1 in AALL07P1, days 1, 4 and 8 of each course in AAML1031.
dose_days_by_study <- list(
  "AALL07P1 (relapsed ALL)" = c(1, 4, 8, 11),
  "AAML1031 (de novo AML)" = c(1, 4, 8)
)
stopifnot(setequal(names(dose_days_by_study), study_tab$treatment))

events <- do.call(bind_rows, lapply(seq_len(nrow(study_tab)), function(i) {
  r <- study_tab[i, ]
  make_cohort(
    n = r$n,
    bsa = draw_bsa(r$n, r$published_mean, r$lo, r$hi),
    study_flag = r$study_flag,
    treatment = r$treatment,
    dose_days = dose_days_by_study[[r$treatment]],
    id_offset = (i - 1L) * 1000L
  )
}))

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))

# The drawn cohort must match the published per-study BSA means. The standard
# error of a 51-subject mean at 35% CV is ~5%, so 15% is a real gate on a
# mis-transcribed mean while admitting sampling noise.
bsa_chk <- events |>
  distinct(id, treatment, BSA) |>
  group_by(treatment) |>
  summarise(
    n_sim = n(), mean_bsa = mean(BSA),
    min_bsa = min(BSA), max_bsa = max(BSA), .groups = "drop"
  ) |>
  left_join(study_tab |> select(treatment, n, published_mean), by = "treatment")
stopifnot(nrow(bsa_chk) == 2L, !anyNA(bsa_chk$published_mean))
knitr::kable(bsa_chk, digits = 3, caption = "Virtual-cohort BSA vs Hanley 2017 Table 1.")
stopifnot(
  # The per-study means are calibrated by construction, so this only confirms
  # the clamp bounds are consistent with them (a range that excluded the mean
  # would still go red).
  all(abs(bsa_chk$mean_bsa / bsa_chk$published_mean - 1) < 0.03),
  identical(bsa_chk$n_sim, bsa_chk$n),
  all(bsa_chk$min_bsa >= 0.60), all(bsa_chk$max_bsa <= 2.53)
)

# Independent cross-check: the pooled mean BSA is a THIRD published number
# (Table 1 total column, 1.30 m^2) that the draw never uses -- it emerges only
# if the per-study means AND the 51/53 split are both transcribed correctly.
pooled_mean_bsa <- mean(distinct(events, id, BSA)$BSA)
stopifnot(
  nrow(distinct(events, id)) == 104L,
  abs(pooled_mean_bsa / 1.30 - 1) < 0.02
)
c(pooled_mean_bsa = pooled_mean_bsa, published_total = 1.30)
```

# Simulation

```{r simulate}
mod <- readModelDb("Hanley_2017_bortezomib")
# `BSA` and `STUDY_AALL07P1` are model covariates and come back automatically;
# only the plot/grouping label needs `keep`. rxSolve returns observation rows
# only and carries no `evid` column, so nothing downstream filters on it.
sim_day8 <- rxode2::rxSolve(mod, events = events, keep = c("treatment")) |>
  as.data.frame()

# `Cc` is the individual prediction; `sim` carries the residual error.
stopifnot(
  all(c("Cc", "sim") %in% names(sim_day8)),
  nrow(sim_day8) > 0,
  all(sim_day8$Cc >= 0)
)
```

# Replicate published figures

## Figure 2 - concentration-time profile after repeat 1.3 mg/m^2 IV dosing

Figure 2 of Hanley 2017 plots the individual day-8 concentrations of all 104
patients over 0-72 h postdose on a log scale, showing the multiexponential
decline that motivated the three-compartment structure. The panel below is the
model's counterpart: the median and 2.5th-97.5th percentile envelope of
simulated concentrations, by study.

```{r figure-2}
prof <- sim_day8 |>
  filter(time >= 168) |>
  mutate(tad = time - 168) |>
  group_by(treatment, tad) |>
  summarise(
    Q025 = quantile(Cc, 0.025),
    Q50 = quantile(Cc, 0.50),
    Q975 = quantile(Cc, 0.975),
    .groups = "drop"
  )

ggplot(prof, aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q025, ymax = Q975), alpha = 0.22) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 0.1, linetype = "dotted") +
  facet_wrap(~treatment) +
  scale_y_log10() +
  labs(
    x = "Time after the day-8 dose (h)", y = "Bortezomib (ng/mL)",
    title = "Figure 2 - day-8 concentration-time profile, 1.3 mg/m2 IV bolus",
    caption = paste(
      "Replicates Figure 2 of Hanley 2017. Dotted line = the 0.1 ng/mL limit of",
      "quantification; the paper reports no concentration below it."
    )
  )
```

The paper reports that none of the 571 observed concentrations were below the
0.1 ng/mL limit of quantification. That is a falsifiable statement about the
model's 72-h trough, so it is asserted rather than merely described.

```{r lloq-check}
trough <- sim_day8 |>
  group_by(id) |>
  summarise(c72 = Cc[which.max(time)], .groups = "drop")
# Median and a robust lower quantile, not the minimum: the per-subject extreme
# of a random cohort is not reproducible across rxode2 builds or thread counts.
stopifnot(
  median(trough$c72) > 0.1,
  quantile(trough$c72, 0.05) > 0.1
)
c(median = median(trough$c72), q05 = quantile(trough$c72, 0.05))
```

## Figure 6 - BSA-normalised clearance is flat across the age range

The paper's central conclusion is that BSA-normalised clearance shows no
discernible relationship with age over 2-16 years (Figure 6, linear-regression
slope 0.0073 L/h/m^2 per year, 95%CI -0.087 to 0.101), which is what supports
BSA-based dosing without an additional age term.

Age is not an input to this model, so the claim cannot be re-tested against an
age axis. What *can* be tested is the mechanism that produces it: with a CL
exponent of 0.97, `CL / BSA` is proportional to `BSA^(-0.03)`, so it is nearly
constant by construction. Across the full published BSA range of
0.60-2.53 m^2 that is a deterministic drift of about 4.4%, which is why no age
trend can survive once BSA is on CL.

```{r figure-6}
mod_typ <- rxode2::zeroRe(mod)
bsa_grid <- seq(0.60, 2.53, length.out = 60)
cl_grid <- 9.59 * (bsa_grid / 1.30)^0.97
drift <- max(cl_grid / bsa_grid) / min(cl_grid / bsa_grid) - 1

# Deterministic: no cohort, no RNG, so a tight bound is correct. CL/BSA is
# proportional to BSA^(exponent - 1), so the drift across the range has the
# closed form below and goes red on a mis-transcribed exponent: 0.75 instead of
# 0.97 would give 39%, and 1.00 would give exactly 0%.
stopifnot(
  abs(drift / ((2.53 / 0.60)^(1 - 0.97) - 1) - 1) < 1e-9,
  drift < 0.05
)

tibble(BSA = bsa_grid, cl_norm = cl_grid / bsa_grid) |>
  ggplot(aes(BSA, cl_norm)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 7.79, linetype = "dashed") +
  expand_limits(y = c(0, 10)) +
  labs(
    x = "BSA (m2)", y = "Typical CL / BSA (L/h/m2)",
    title = "Figure 6 mechanism - CL/BSA is flat because the BSA exponent is 0.97",
    caption = paste0(
      "Total drift across the published BSA range 0.60-2.53 m2 is ",
      round(100 * drift, 1), "%. Dashed line = the paper's reported ",
      "all-patient geometric mean of 7.79 L/h/m2."
    )
  )
```

# Structural gates

Before comparing against the paper's numbers, two checks confirm the encoded
system is the system that was intended. Both compare a solve against its own
closed form, so the difference is pure numerical error and a tight bound is
correct.

`rxode2` can silently replace an explicit ODE system with an analytic
`linCmt()` solution when it recognises a `cl` / `vc` parameter pair, which would
quietly discard the ODE block. It does not fire here, and the first assertion
records that.

```{r structural-gates}
stopifnot(
  is.null(ui$linCmt),
  identical(ui$state, c("central", "peripheral1", "peripheral2"))
)

# Mass balance: for a linear system, cl * AUCinf must recover the whole dose.
# Cc is in ng/mL and the dose in mg, hence the factor 1000.
dose_ref <- 1.3 * 1.30
ev_long <- rxode2::et(amt = dose_ref, cmt = "central") |>
  rxode2::et(seq(0, 1000, length.out = 20001))
solve_ref <- function(bsa) {
  rxode2::rxSolve(
    mod_typ, ev_long,
    params = c(BSA = bsa, STUDY_AALL07P1 = 0),
    omega = NA, returnType = "data.frame", atol = 1e-12, rtol = 1e-10
  )
}
s_ref <- solve_ref(1.30)
stopifnot(all(s_ref$Cc >= 0))
auc_ref <- PKNCA::pk.calc.auc.last(conc = s_ref$Cc, time = s_ref$time)
cl_recovered <- dose_ref / (auc_ref / 1000)
stopifnot(abs(cl_recovered / 9.59 - 1) < 0.01)

# The BSA power on CL must be the exponent in ini(), recovered end-to-end from
# the solve rather than read off the parameter.
s_2x <- solve_ref(2.60)
cl_2x <- dose_ref / (PKNCA::pk.calc.auc.last(conc = s_2x$Cc, time = s_2x$time) / 1000)
stopifnot(abs((cl_2x / cl_recovered) / 2^0.97 - 1) < 0.01)

c(
  cl_recovered = cl_recovered,
  exponent_recovered = log(cl_2x / cl_recovered) / log(2)
)
```

# PKNCA validation

The paper publishes no NCA table, so PKNCA is used to recover the disposition
quantities the paper *does* report as individual-parameter summaries. Because
the model is linear and time-invariant, clearance, steady-state volume and the
three disposition half-lives are properties of the system rather than of a
particular schedule, so they are computed from a single-dose profile followed
to 1000 h (ten terminal half-lives).

```{r pknca}
subj <- events |> distinct(id, treatment, BSA, STUDY_AALL07P1)

ev_nca <- bind_rows(
  subj |> mutate(time = 0, amt = 1.3 * BSA, evid = 1L, cmt = "central"),
  subj |> tidyr::crossing(time = c(
    seq(0, 2, length.out = 41), seq(2.5, 24, by = 0.5), seq(26, 1000, by = 6)
  )) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
) |>
  arrange(id, time, desc(evid))

sim_nca_raw <- rxode2::rxSolve(mod, events = ev_nca, keep = c("treatment")) |>
  as.data.frame()

# Only `!is.na(Cc)`: a `time > 0` or `Cc > 0` filter would drop the time-zero
# row PKNCA needs to anchor AUC0-*.
sim_nca <- sim_nca_raw |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)
stopifnot(nrow(sim_nca) > 0, all(sim_nca$Cc >= 0))

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- ev_nca |>
  filter(evid == 1L) |>
  select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE,
  aumcinf.obs = TRUE, mrt.obs = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  select(treatment, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  left_join(subj |> select(id, BSA), by = "id") |>
  mutate(
    # Cc is ng/mL and amt is mg, so AUC is ng*h/mL = ug*h/L and CL in L/h needs
    # the dose in ug.
    cl_nca = (1.3 * BSA * 1000) / aucinf.obs,
    vss_nca = cl_nca * mrt.obs,
    cl_per_bsa = cl_nca / BSA,
    vss_per_bsa = vss_nca / BSA
  )
stopifnot(nrow(nca_wide) == 104L, !anyNA(nca_wide$aucinf.obs))
```

## Comparison against the published terminal half-life

The terminal (gamma) disposition half-life is the one published quantity that
is a bona fide NCA parameter, so it goes through
`nlmixr2lib::ncaComparisonTable()`. Hanley 2017 reports a single all-patient
geometric mean of 100 h; using it as the reference for both study groups also
tests the paper's separate claim that bortezomib PK did not differ between the
ALL and AML populations.

```{r nca-compare}
# One reference value for both groups; derived from study_tab so the labels
# cannot drift apart from the cohort's.
published <- study_tab |>
  select(treatment) |>
  mutate(half.life = 100)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "treatment",
  units = c(half.life = "h"),
  tolerance_pct = 20
)

cmp |>
  knitr::kable(
    caption = paste(
      "Simulated vs published terminal half-life (Hanley 2017 Results:",
      "100 h, 44 %CV). * differs from reference by >20%."
    )
  )
```

## Comparison against the published individual-parameter summary

The remaining published quantities are summaries of the model's individual
parameters rather than NCA output (Hanley 2017 Results, final paragraph before
the Discussion): a geometric mean BSA-normalised clearance of 7.79 L/(h m^2)
with 25 %CV, a steady-state volume of 834 L/m^2 with 39 %CV, and alpha, beta
and gamma disposition half-lives of 6.15 min, 1.58 h and 100 h. The alpha and
beta half-lives are not recoverable by NCA on a sparse grid, so they are
computed as eigenvalues of the per-subject rate-constant matrix, using each
subject's own `cl`, `vc`, `q`, `vp`, `q2` and `vp2` as returned by the solve.

```{r disposition-halflives}
pars <- sim_nca_raw |>
  group_by(id) |>
  slice(1) |>
  ungroup() |>
  select(id, treatment, BSA, cl, vc, q, vp, q2, vp2)
stopifnot(nrow(pars) == 104L)

tri_halflives <- function(cl, vc, q, vp, q2, vp2) {
  k10 <- cl / vc
  k12 <- q / vc
  k21 <- q / vp
  k13 <- q2 / vc
  k31 <- q2 / vp2
  a2 <- k10 + k12 + k21 + k13 + k31
  a1 <- k10 * k21 + k10 * k31 + k12 * k31 + k13 * k21 + k21 * k31
  a0 <- k10 * k21 * k31
  lam <- sort(Re(polyroot(c(-a0, a1, -a2, 1))), decreasing = TRUE)
  log(2) / lam
}

hl <- t(mapply(tri_halflives, pars$cl, pars$vc, pars$q, pars$vp, pars$q2, pars$vp2))
pars$t_alpha_min <- hl[, 1] * 60
pars$t_beta_h <- hl[, 2]
pars$t_gamma_h <- hl[, 3]

gm <- function(x) exp(mean(log(x)))
cv_pct <- function(x) 100 * stats::sd(x) / mean(x)

derived <- tibble::tribble(
  ~quantity,                            ~published, ~published_cv, ~simulated,                      ~sim_cv,                       ~route,
  "CL / BSA (L/[h m2])",                7.79,       25,            gm(nca_wide$cl_per_bsa),         cv_pct(nca_wide$cl_per_bsa),   "PKNCA AUCinf",
  "Vss / BSA (L/m2)",                   834,        39,            gm(nca_wide$vss_per_bsa),        cv_pct(nca_wide$vss_per_bsa),  "PKNCA AUCinf x MRT",
  "alpha half-life (min)",              6.15,       32,            gm(pars$t_alpha_min),            cv_pct(pars$t_alpha_min),      "eigenvalue",
  "beta half-life (h)",                 1.58,       18,            gm(pars$t_beta_h),               cv_pct(pars$t_beta_h),         "eigenvalue",
  "gamma half-life (h)",                100,        44,            gm(pars$t_gamma_h),              cv_pct(pars$t_gamma_h),        "eigenvalue"
) |>
  mutate(pct_diff = 100 * (simulated / published - 1))

derived |>
  rename(
    "Quantity" = quantity,
    "Published" = published,
    "Published %CV" = published_cv,
    "Simulated (geometric mean)" = simulated,
    "Simulated %CV" = sim_cv,
    "Derivation" = route,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(
    digits = c(0, 2, 0, 2, 0, 0, 1),
    caption = "Simulated vs published individual-parameter summary (Hanley 2017 Results)."
  )
```

All five quantities agree with the paper within a common, one-sided offset of
roughly 5%, which is the residual of the BSA-normalisation choice discussed
below rather than a transcription error: it moves every clearance-derived
quantity in the same direction by the same fraction, whereas a mis-transcribed
volume, dose or unit would move individual rows by tens of percent.

```{r derived-gate}
# Assertions on cohort geometric means, so the bound must admit the eta draw.
# Measured at 1, 2 and 16 solver threads: -6.5 / -1.0 / -4.5 / +0.2 /
# +5.7 percent on the five rows, IDENTICAL at all three thread counts (the BSA
# draw is R's RNG, which is thread-independent, and a 104-subject geometric mean
# is insensitive to the eta partitioning). Worst observed magnitude 6.5%, and
# the systematic part of that is the ~5% normalisation offset discussed below.
# Do NOT tighten below ~12%: the bound must survive a different cohort. 20% is
# the skill's standard tolerance and still goes red on a mis-transcribed value
# (a volume or dose error moves a row by tens of percent; reading the BSA
# reference as 1.8 m2 moves the gamma half-life by +40%).
stopifnot(
  all(abs(derived$pct_diff) < 20),
  # Structural: the centre of the distribution must not drift, which is the
  # sharpest single statement about the transcription as a whole.
  abs(median(derived$pct_diff)) < 10
)
round(derived$pct_diff, 2)
```

## Study effect is residual-only

The retained `STUDY_AALL07P1` covariate selects a residual-error magnitude and
nothing else: disease type was screened on the structural parameters and was
not significant. Two consequences are checked. First, the individual
predictions for a subject must not depend on the study flag at all. Second, the
simulated observations must, because the residual magnitude differs by roughly
a factor of two (46.8 %CV vs 21.9 %CV).

```{r study-effect}
ev_one <- rxode2::et(amt = 1.3 * 1.30, cmt = "central") |>
  rxode2::et(seq(0.5, 72, by = 0.5))
ipred_flag <- vapply(c(0, 1), function(f) {
  s <- rxode2::rxSolve(
    mod_typ, ev_one, params = c(BSA = 1.30, STUDY_AALL07P1 = f),
    omega = NA, returnType = "data.frame"
  )
  max(s$Cc)
}, numeric(1))
# Deterministic comparison of two typical-value solves: exact equality is the
# correct assertion here, not a tolerance.
stopifnot(isTRUE(all.equal(ipred_flag[1], ipred_flag[2], tolerance = 1e-10)))

# Encoded SDs joined BY NAME via the study flag, never positionally.
encoded_sd <- study_tab |>
  select(treatment, study_flag) |>
  mutate(encoded = sqrt(log(1 + ifelse(study_flag == 1, 0.468, 0.219)^2))) |>
  select(treatment, encoded)

resid_sd <- sim_day8 |>
  filter(Cc > 0, sim > 0) |>
  group_by(treatment) |>
  summarise(n_rec = n(), sd_log_resid = stats::sd(log(sim / Cc)), .groups = "drop") |>
  left_join(encoded_sd, by = "treatment")
stopifnot(nrow(resid_sd) == 2L, !anyNA(resid_sd$encoded), all(resid_sd$n_rec > 1000))
knitr::kable(resid_sd, digits = 4, caption = "Realised vs encoded log-scale residual SD, by study.")
# Sampling SD of an SD over ~7500 records per arm is well under 2%, so 15% is
# generous while still catching a swapped or dropped study-specific sigma
# (swapping them would put each arm ~2x off).
stopifnot(all(abs(resid_sd$sd_log_resid / resid_sd$encoded - 1) < 0.15))
```

# Assumptions and deviations

## The BSA normalisation constant: the paper contradicts itself, and its own numbers settle it

This is the one load-bearing judgement in the extraction, so the reasoning is
recorded in full.

Hanley 2017's Methods give the covariate model as a power function and define
its terms as follows: "`covi` represents the individual value for the covariate
**normalized for the population mean**". The very next sentence then says: "All
body size metrics were referenced to the **size of an average adult** (weight,
70 kg; BSA, 1.8 m2)." For a cohort whose mean BSA is 1.30 m^2, these two
statements differ by a factor of `(1.8 / 1.30)^0.97 = 1.37` on every clearance
the model predicts, so the choice cannot be deferred.

The paper's own reported derived quantities refute the 1.8 m^2 reading and are
consistent with the population-mean reading. Reading the constant as 1.8 m^2
means the tabulated CL of 9.59 L/h belongs to a 1.8 m^2 adult, so a typical
patient in this cohort would have a much lower clearance, and:

| Published quantity | Reported | Predicted if reference = 1.8 m^2 | Predicted if reference = the population mean |
|---|---|---|---|
| alpha half-life | 6.15 min | 7.36 min (+20%) | 6.50 min (+6%) |
| beta half-life | 1.58 h | 1.82 h (+15%) | 1.59 h (+0%) |
| gamma half-life | 100 h | 140 h (+40%) | 104 h (+4%) |
| CL / BSA | 7.79 L/(h m2) | 5.39 L/(h m2) (-31%) | 7.39 L/(h m2) (-5%) |

Two further internal checks corroborate the population-mean reading
independently of the half-lives. First, the reported steady-state volume of
834 L/m^2 divided into `V1 + V2 + V3 = 1017.5 L` implies a cohort geometric-mean
BSA of 1.22 m^2 - and it does so only because no volume carries a BSA effect,
which is exactly what Table 2 shows. Second, inverting the reported
`CL / BSA = 7.79 L/(h m2)` for the normalisation constant gives 1.23 m^2, which
agrees with that independently derived geometric mean to within 1%.

**What is encoded, and why.** The model uses **1.30 m^2**, the total-cohort mean
BSA printed in Table 1. That is a published number, which the back-solved
1.23 m^2 is not, and the library's rule is that every `ini()` value trace to the
source. The cost is a systematic offset of about 5% on every clearance-derived
quantity, visible as the common one-sided bias in the comparison table above,
and arising because the paper appears to have normalised by a central value
nearer the cohort's geometric mean or median (~1.23 m^2) than its arithmetic
mean (1.30 m^2; the median BSA is not printed). Anyone needing the paper's
derived summaries reproduced to within 1% rather than 5% should drive the model
with `BSA / 1.30 * 1.23` in place of `BSA`; the structure and every other
parameter are unaffected.

## Other assumptions

- **BSA distribution.** Table 1 reports per-study means and ranges but no SD or
  percentiles, so BSA is drawn log-normally at 35 %CV - the value implied by the
  published mean and range - and clamped to the reported range. The BSA formula
  (DuBois, Mosteller or Haycock) is not stated in the paper, so the cohort
  samples BSA directly from the reported distribution rather than recomputing it
  from height and weight.
- **%CV to variance conversion.** Table 2 reports both the IIV and the residual
  error as percentage coefficients of variation. Both are on the log scale in
  this model (exponential IIV; log-transform-both-sides residual), so both are
  converted with the exact log-normal relation `omega^2 = log(CV^2 + 1)`,
  giving residual SDs of 0.445 and 0.216. The looser convention of taking
  `sigma = CV` directly would give 0.468 and 0.219 - about 5% higher on the
  larger of the two - and neither reading affects any structural gate above.
- **Residual error encoded as `lnorm`, not `prop`.** The paper states it used
  the log-transform-both-sides approach with a homoscedastic log-scale error,
  which is `lnorm()` in nlmixr2, and separately glosses it as "proportional
  after back transformation". The mechanism, not the gloss, is encoded, matching
  the `Friberg_2012_voriconazole` precedent for the same wording.
- **Concentration units.** `central` is an amount in mg and `vc` a volume in L,
  so `central / vc` is mg/L. The model multiplies by 1000 to report ng/mL, the
  unit of the paper's LC-MS/MS assay (dynamic range 0.1-25.0 ng/mL). The paper
  gives no scaling factor, so this is inferred from the assay units.
- **Screened-but-unretained covariates.** Age, weight, race, disease type, AML
  risk group and ALL treatment-plan stratum were all screened and none reached
  significance, so none has a published point estimate. They are documented in
  the model's `covariatesDataExcluded` metadata rather than `covariateData`.
- **`STUDY_AALL07P1` carries no structural effect.** It is retained solely to
  select the residual-error magnitude, which is why the study-effect check above
  asserts that individual predictions are identical across its two levels.
- **Repeat-dose scope.** The model must not be used for first-dose exposure.
  The authors state that bortezomib clearance falls between the first dose of
  cycle 1 and steady state and that this model, fit to day-8 repeat-dose data,
  "cannot be directly used to describe systemic exposures following the first
  bortezomib dose". Extrapolation below 2 years of age is likewise outside the
  studied range.
- **Figures not reproduced.** Figures 3 (goodness-of-fit), 4 (covariate-vs-eta
  plots) and 5 (VPC) are diagnostics of the original fit against the original
  observations, which are not public, so they have no simulation counterpart.
  Figure 6's age axis is not a model input; the mechanism that produces its flat
  result is tested instead, as described in that section.
- **No erratum found.** A search of the publisher's record for this DOI and of
  PubMed returned no correction, corrigendum or erratum.
