---
title: "Aripiprazole lauroxil (Hard 2017)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aripiprazole lauroxil (Hard 2017)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
ui <- rxode2::rxode(readModelDb("Hard_2017_aripiprazole_lauroxil"))
```

* Citation: `r ui$reference`
* Description: `r ui$description`
* Article: <https://doi.org/10.1007/s40263-017-0447-7>
* Electronic Supplementary Material (parameter table): <https://doi.org/10.1007/s40263-017-0447-7> (Supplementary file 1, `40263_2017_447_MOESM1_ESM.pdf`)
* Earlier PopPK model from which Vp/F, Q/F and the weight centering are carried:
  <https://doi.org/10.1097/JCP.0000000000000685>

Aripiprazole lauroxil (AL) is an intramuscular extended-release prodrug of
aripiprazole. The prodrug itself is not measurable in plasma; the model
therefore describes aripiprazole only, entering the systemic circulation as a
**lagged zero-order input** whose duration (43.5 days) exceeds every approved
dosing interval. The paper calls this model the **2MPopPK model**.

## Structure

Supplemental Fig. 1 of the ESM draws seven intramuscular depots and one oral
depot all feeding a single aripiprazole central compartment, which exchanges
with one peripheral compartment and is cleared by CL/F. The seven IM depots are
a NONMEM book-keeping device -- the figure's own footnote says "a new IM depot
was added for each injection ... to accommodate the maximum number of AL
injections in the clinical studies (seven in study A105)" -- and the model
carries **no rate constant out of an IM depot**. Each depot contributes a
zero-order input of duration `D1` beginning `ALAG` after the injection, so the
seven depots superpose exactly onto a single lagged modelled-duration input to
`central`. That is how the model is encoded here, which also removes the
artificial seven-injection ceiling.

```{r structure}
cat(paste(rxode2::modelExtract(ui), collapse = "\n"))
```

Consequences for a user's data set:

* **IM aripiprazole lauroxil doses** use `cmt = "central"` with `rate = -2`
  (so rxode2 uses the modelled duration `dur(central) <- d1`) and the
  aripiprazole-**equivalent** dose in mg, not the AL dose. The paper's mapping
  is 441 / 662 / 882 / 1064 mg AL = 300 / 450 / 600 / 724 mg aripiprazole
  equivalent (Sect. 2.2). Without `rate = -2` rxode2 silently delivers a bolus.
* **Oral aripiprazole doses** use `cmt = "depot"` and are the FPO = 1 reference
  route.
* `OCC` carries the injection number (1..7) on each IM dose record and
  multiplexes the inter-occasion variability on `D1`; `OCC = 0` switches it off.

# Population

```{r population}
pop <- ui$population
str(pop, max.level = 1)
```

The 2MPopPK data set pooled **14,524 aripiprazole concentrations from 700
patients** with schizophrenia or schizoaffective disorder across five studies:
the phase I study A105 (NCT02320032; 124 patients, 5077 concentrations), three
earlier phase I studies (002, 101, 102) and the pivotal phase III study 003
(ESM Supplemental Table 2). 826 concentrations (6%) were below the limit of
quantification and were retained using the M3 method. Study 001 of the earlier
PopPK analysis was dropped because it used a non-commercial formulation.

Demographics quoted in the model metadata are the A105 safety population of
Table 1 (mean age 44.5 years, 73% male, 75% Black or African American, mean
weight 89.8 kg). CYP2D6 phenotype in the pooled data set was 25 poor, 416
extensive, 183 intermediate, 3 inconclusive and 73 missing; **no ultra-rapid
metabolizers** were present, so the model offers no guidance for that phenotype.

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Hard_2017_aripiprazole_lauroxil.R` carries an
in-file comment pointing at its source location. They are collected here.

| Parameter / equation | Value | Source location |
|---|---|---|
| `lka` (Ka) | 0.803 1/h | ESM Supplemental Table 7, row `Ka (h-1)` (%RSE 29.9) |
| `lcl` (CL/F) | 1.898 L/h | ESM Supplemental Table 7, row `CL/F (L/hr)` (%RSE 2.57) |
| `lvc` (VC/F) | 317 L | ESM Supplemental Table 7, row `VC/F (L)` (%RSE 2.25) |
| `lvp` (VP/F) | 2122 L, fixed | ESM Supplemental Table 7, row `VP/F (L)`, `**` + footnote a |
| `lq` (Q/F) | 0.423 L/h, fixed | ESM Supplemental Table 7, row `Q/F (L/hr)`, `**` + footnote a |
| `ld1` (D1) | 1043 h | ESM Supplemental Table 7, row `D1 (hr)` (%RSE 2.09) |
| `ltlag` (ALAG) | 77.5 h | ESM Supplemental Table 7, row `ALAG (hr)` (%RSE 4.07) |
| `lc0` (ARI(0)) | 0.915 ng/mL, fixed | ESM Supplemental Table 7, row `ARI(0) (ng/mL)`, `**`; ESM Sect. 4 |
| `lfdepot_im` (FIM) | 0.571 | ESM Supplemental Table 7, row `FIM` (%RSE 2.55) |
| `lfdepot` (FPO) | 1, fixed | ESM Supplemental Table 7, row `FPO`, `**` + footnote b |
| `e_wt_vc` | 1, fixed | ESM Supplemental Table 7, row `WT ON VC/F`, `**` + footnote b |
| weight centering (70 kg) | 70 kg | Hard 2017 *J Clin Psychopharmacol* Supplementary Data Content: "centered for a weight of 70 kg"; `VC/F = 268*(WT/70)^1.0` |
| `e_cyp2d6_pm_cl` | 0.767 | ESM Supplemental Table 7, row `CL/F PMs` (%RSE 10.2) + footnote c |
| IIV variances (9 etas) | see model file | ESM Supplemental Table 7, "Inter-individual Variability" Value column |
| IOV on `D1` | 0.125 | ESM Supplemental Table 7, "Inter-occasion Variability" block |
| `propSdStdyA105` | sqrt(0.0238) | ESM Supplemental Table 7, `sigma2prop ARI A105` |
| `propSdStdyPrior` | sqrt(0.0565) | ESM Supplemental Table 7, `sigma2prop ARI Non-A105` |
| Model topology (7 IM depots + oral depot -> central <-> peripheral) | n/a | ESM Supplemental Fig. 1 |
| Lagged zero-order IM input | n/a | ESM Supplemental Fig. 1 footnote; Sect. 3.2 ("43 days ... 3.2-day lag") |

## The IIV column is a variance

The "Inter-individual Variability / Value" column holds `omega^2` on the log
scale. Footnote `d` of Supplemental Table 7 gives
`CV = sqrt(exp(omega^2) - 1) * 100` for the rows it flags, and the rows it does
*not* flag use the small-variance approximation `sqrt(omega^2)` that ESM
Sect. 1 describes for variances below 0.15. Both readings are reproduced below
from the model file's own values, which confirms the variance interpretation.

```{r iiv-check}
iniDf <- ui$iniDf
etas <- iniDf |>
  dplyr::filter(!is.na(neta1), neta1 == neta2, grepl("^etal", name)) |>
  dplyr::transmute(
    eta = name,
    omega2 = est,
    `CV% via sqrt(exp(w2)-1)` = round(100 * sqrt(exp(est) - 1), 1),
    `CV% via sqrt(w2)` = round(100 * sqrt(est), 1)
  )
published_cv <- c(
  etalka = 367, etalcl = 66.5, etalvc = 44.7, etalvp = 435, etalq = 145,
  etald1 = 61.2, etaltlag = 64.1, etalc0 = 1898, etalfdepot_im = 32.2
)
etas$`Published CV%` <- published_cv[etas$eta]
knitr::kable(etas, digits = 3,
             caption = "Supplemental Table 7 IIV column reproduced from the encoded variances. Every row matches one of the two formulae the table itself uses; FIM (32.2%) is the row that uses the small-variance approximation.")

# Each published CV% must be reproduced by one of the two formulae to within
# rounding. This is arithmetic on fixed numbers -- no simulation, so the bound
# is tight by construction.
matched <- pmin(
  abs(etas$`CV% via sqrt(exp(w2)-1)` - etas$`Published CV%`),
  abs(etas$`CV% via sqrt(w2)` - etas$`Published CV%`)
)
stopifnot(length(matched) == 9L, all(matched < 1.0))
```

# Structural checks

These are deterministic: they solve the typical-value model (`zeroRe()`) and
compare against closed-form identities, so their tolerances are numerical, not
statistical.

```{r typical-model}
mod_tv <- rxode2::zeroRe(ui)

th <- setNames(ui$theta, names(ui$theta))
cl_typ <- exp(th[["lcl"]])
fim <- exp(th[["lfdepot_im"]])
d1_typ <- exp(th[["ld1"]])
tlag_typ <- exp(th[["ltlag"]])
c0_typ <- exp(th[["lc0"]])
vc70 <- exp(th[["lvc"]])
vp_typ <- exp(th[["lvp"]])
q_typ <- exp(th[["lq"]])

solve_tv <- function(events) {
  suppressWarnings(rxode2::rxSolve(mod_tv, events, addDosing = FALSE)) |>
    as.data.frame() |>
    dplyr::filter(!is.na(Cc))
}
trap <- function(tm, cc) sum(diff(tm) * (head(cc, -1) + tail(cc, -1)) / 2)
```

## Mass balance: `CL/F * AUC(0-inf) == FIM * Dose + A0`

The single invariant that a mis-transcribed clearance, dose, bioavailability or
unit conversion cannot survive. It holds separately for each input route, and it
is blind to how the dose is split in time -- so it is run on both the IM and the
oral path.

```{r mass-balance}
# The oral route peaks within hours while the IM route peaks after six weeks,
# so the observation grid is dense over the first two days and coarse
# afterwards. A grid that does not resolve Tmax understates AUC by several
# percent and would make the mass-balance identity below look violated when it
# is not.
mb_events <- function(cmt, amt, times_h, occ, horizon_h) {
  d <- data.frame(id = 1L, time = times_h, amt = amt, evid = 1L, cmt = cmt,
                  rate = if (cmt == "central") -2 else NA_real_, OCC = occ,
                  WT = 70, CYP2D6_PM = 0, STUDY_A105 = 1)
  grid <- sort(unique(c(seq(0, 48, by = 0.25), seq(0, horizon_h, by = 6))))
  o <- data.frame(id = 1L, time = grid, amt = NA_real_,
                  evid = 0L, cmt = "central", rate = NA_real_, OCC = 0,
                  WT = 70, CYP2D6_PM = 0, STUDY_A105 = 1)
  dplyr::arrange(dplyr::bind_rows(d, o), time, dplyr::desc(evid))
}

# Seven overlapping IM injections; each 46-day input overlaps its neighbours.
im <- solve_tv(mb_events("central", 300, 672 * (0:6), 1:7, 80000))
im_recovered <- cl_typ * trap(im$time, im$Cc) / 1000
im_expected <- fim * 7 * 300 + c0_typ * vc70 / 1000

# Single oral dose through the FPO = 1 depot.
po <- solve_tv(mb_events("depot", 15, 0, 0, 80000))
po_recovered <- cl_typ * trap(po$time, po$Cc) / 1000
po_expected <- 1 * 15 + c0_typ * vc70 / 1000

mass_balance <- tibble::tibble(
  Route = c("IM, 7 x 300 mg q4wk (overlapping inputs)", "Oral, single 15 mg"),
  `Recovered CL/F * AUCinf (mg)` = c(im_recovered, po_recovered),
  `Expected F * Dose + A0 (mg)` = c(im_expected, po_expected),
  Ratio = c(im_recovered / im_expected, po_recovered / po_expected)
)
knitr::kable(mass_balance, digits = c(0, 3, 3, 5))

# Pure numerical error (grid truncation of a tail that is still decaying), so a
# tight bound is correct here -- both sides use the same drawn parameters.
stopifnot(all(abs(mass_balance$Ratio - 1) < 0.005))
```

## The zero-order input window

The single-dose peak must sit exactly at `ALAG + D1`, the moment the input
stops, and nothing may enter the central compartment during the lag.

```{r input-window}
sd_dose <- data.frame(id = 1L, time = 0, amt = 300, evid = 1L, cmt = "central",
                      rate = -2, OCC = 1, WT = 70, CYP2D6_PM = 0, STUDY_A105 = 1)
sd_obs <- data.frame(id = 1L, time = seq(0, 6000, by = 1), amt = NA_real_,
                     evid = 0L, cmt = "central", rate = NA_real_, OCC = 0,
                     WT = 70, CYP2D6_PM = 0, STUDY_A105 = 1)
sd <- solve_tv(dplyr::arrange(dplyr::bind_rows(sd_dose, sd_obs), time,
                              dplyr::desc(evid)))

tmax_obs <- sd$time[which.max(sd$Cc)]
# During the lag only the baseline decays; the very next sample after the lag
# ends must already be rising.
pre_lag <- sd$Cc[sd$time == floor(tlag_typ) - 7]
post_lag <- sd$Cc[sd$time == ceiling(tlag_typ) + 7]

input_window <- tibble::tibble(
  Quantity = c("Tmax after a single IM dose (h)", "ALAG + D1 (h)",
               "Cc 7 h before the lag ends (ng/mL)", "Cc 7 h after the lag ends (ng/mL)"),
  Value = c(tmax_obs, tlag_typ + d1_typ, pre_lag, post_lag)
)
knitr::kable(input_window, digits = 3)

stopifnot(
  abs(tmax_obs - (tlag_typ + d1_typ)) <= 1, # grid resolution is 1 h
  pre_lag < c0_typ, # still only the decaying baseline
  post_lag > pre_lag # input has started
)
```

## Covariate effects

The CYP2D6 poor-metabolizer effect is a pure clearance multiplier, so the
exposure ratio must be exactly `1 / 0.767`. The weight effect is an exponent of
1 on Vc/F centred at 70 kg, so it must not move exposure at all.

```{r covariate-checks}
auc_for <- function(pm, wt) {
  ev <- mb_events("central", 300, 672 * (0:6), 1:7, 80000)
  ev$CYP2D6_PM <- pm
  ev$WT <- wt
  s <- solve_tv(ev)
  trap(s$time, s$Cc)
}
auc_ref <- auc_for(0, 70)
cov_checks <- tibble::tibble(
  Check = c("AUCinf ratio, CYP2D6 PM vs non-PM", "AUCinf ratio, 120 kg vs 70 kg"),
  Simulated = c(auc_for(1, 70) / auc_ref, auc_for(0, 120) / auc_ref),
  Expected = c(1 / exp(log(ui$theta[["e_cyp2d6_pm_cl"]])), 1),
  Note = c("CL/F multiplier 0.767 (ESM Suppl. Table 7)",
           "WT acts on Vc/F only; exposure is CL-determined")
)
knitr::kable(cov_checks, digits = 5)
stopifnot(all(abs(cov_checks$Simulated - cov_checks$Expected) < 1e-3))
```

## Disposition half-life implied by the model

Worth computing explicitly because it is where the model and the paper's own
non-compartmental analysis part company (see *Deviations*). With Vp/F and Q/F
fixed at the earlier model's estimates the terminal (beta) phase is far slower
than the 54-57 day half-lives the phase I study measured.

```{r beta-halflife}
k10 <- cl_typ / vc70
k12 <- q_typ / vc70
k21 <- q_typ / vp_typ
sum_ab <- k10 + k12 + k21
beta <- (sum_ab - sqrt(sum_ab^2 - 4 * k10 * k21)) / 2
alpha <- (sum_ab + sqrt(sum_ab^2 - 4 * k10 * k21)) / 2
t_half_beta <- log(2) / beta / 24
t_half_alpha <- log(2) / alpha / 24
tibble::tibble(
  Phase = c("alpha (distribution)", "beta (terminal)"),
  `Half-life (days)` = c(t_half_alpha, t_half_beta)
) |>
  knitr::kable(digits = 2,
               caption = "Two-compartment eigenvalue half-lives at 70 kg, from the model's own CL/F, Vc/F, Q/F and Vp/F.")
```

# Replicating the phase I study (A105)

Study A105 gave four, five or seven injections with the **last dose on day 169**
for every arm, then followed patients for a further 20 weeks. Table 2 of the
paper reports non-compartmental parameters over the interval **following that
last dose**, so the simulation below reproduces the trial exactly: the same
injection days, the same 48 scheduled PK sampling days (Supplemental Table 2),
no oral aripiprazole, and the aripiprazole-equivalent doses.

```{r a105-design}
# ESM Supplemental Table 2, study A105 PK sampling days.
sampling_day <- c(1, 4, 8, 11, 15, 18, 22, 25, 29, 32, 36, 39, 43, 46, 50, 53,
                  57, 71, 85, 99, 113, 127, 141, 155, 169, 172, 176, 180, 183,
                  186, 190, 193, 197, 200, 204, 207, 211, 214, 217, 221, 225,
                  228, 232, 239, 246, 253, 281, 309)

arms <- list(
  list(regimen = "441 q4wk",  eq = 300, dose_day = c(1, 29, 57, 85, 113, 141, 169), tau_d = 28),
  list(regimen = "882 q6wk",  eq = 600, dose_day = c(1, 43, 85, 127, 169),          tau_d = 42),
  list(regimen = "1064 q8wk", eq = 724, dose_day = c(1, 57, 113, 169),              tau_d = 56)
)
last_dose_day <- 169
followup_day <- 140 # Table 2 footnote a: duration of sampling, 140 days

make_arm <- function(a, ids, wt) {
  dose <- expand.grid(k = seq_along(ids), j = seq_along(a$dose_day))
  dose <- data.frame(
    id = ids[dose$k], time = (a$dose_day[dose$j] - 1) * 24, amt = a$eq,
    evid = 1L, cmt = "central", rate = -2, OCC = dose$j,
    WT = wt[dose$k], CYP2D6_PM = 0, STUDY_A105 = 1, regimen = a$regimen
  )
  obs <- expand.grid(k = seq_along(ids), t = (sampling_day - 1) * 24)
  obs <- data.frame(
    id = ids[obs$k], time = obs$t, amt = NA_real_, evid = 0L, cmt = "central",
    rate = NA_real_, OCC = 0, WT = wt[obs$k], CYP2D6_PM = 0, STUDY_A105 = 1,
    regimen = a$regimen
  )
  dplyr::arrange(dplyr::bind_rows(dose, obs), id, time, dplyr::desc(evid))
}
```

## Typical-value replication

```{r a105-typical}
ev_tv <- dplyr::bind_rows(lapply(seq_along(arms), function(i) {
  make_arm(arms[[i]], ids = i, wt = 89.8) # A105 mean weight, Table 1
}))
sim_tv <- suppressWarnings(
  rxode2::rxSolve(mod_tv, ev_tv, keep = c("regimen"), addDosing = FALSE)
) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc))

ggplot(sim_tv, aes(time / 24, Cc, colour = regimen)) +
  geom_line() +
  geom_point(size = 0.9) +
  geom_vline(xintercept = last_dose_day - 1, linetype = 2, colour = "grey40") +
  labs(x = "Study day", y = "Aripiprazole (ng/mL)", colour = "Regimen",
       title = "Typical-value replication of the A105 design",
       caption = paste("Compare with Figure 1 of Hard 2017. Dashed line = last",
                       "injection (day 169). Sampling days are the 48 scheduled",
                       "A105 PK days of ESM Supplemental Table 2."))
```

## PKNCA over the interval following the last dose

The NCA frame carries time in **days** so the output is directly comparable
with Table 2, which reports AUC in day*ng/mL and half-life in days; the model's
native time unit is hours.

```{r pknca}
nca_conc <- sim_tv |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(time_d = time / 24) |>
  dplyr::select(id, time_d, Cc, regimen)

nca_dose <- ev_tv |>
  dplyr::filter(evid == 1) |>
  dplyr::mutate(time_d = time / 24) |>
  dplyr::select(id, time_d, amt, regimen)

# Two windows per arm: the dosing interval after the last injection (Cmax,
# Tmax, Cavg, AUCtau) and the full 140-day follow-up (half-life).
intervals <- do.call(rbind, lapply(arms, function(a) {
  data.frame(
    regimen = a$regimen,
    start = c(last_dose_day - 1, last_dose_day - 1),
    end = c(last_dose_day - 1 + a$tau_d, last_dose_day - 1 + followup_day),
    cmax = c(TRUE, FALSE), tmax = c(TRUE, FALSE), cav = c(TRUE, FALSE),
    auclast = c(TRUE, FALSE), half.life = c(FALSE, TRUE)
  )
}))

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ time_d | regimen + id,
                             concu = "ng/mL", timeu = "day")
dose_obj <- PKNCA::PKNCAdose(nca_dose, amt ~ time_d | regimen + id,
                             doseu = "mg")
nca_res <- suppressWarnings(
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
)
```

## Comparison against the published non-compartmental analysis

```{r nca-compare}
# Hard 2017 Table 2, arithmetic means following the last dose.
published <- tibble::tribble(
  ~regimen,     ~cmax, ~cav, ~auclast, ~half.life,
  "441 q4wk",   161,   126,  3520,     57.2,
  "882 q6wk",   172,   131,  5510,     55.1,
  "1064 q8wk",  189,   141,  7880,     53.9
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "regimen",
  units = c(cmax = "ng/mL", cav = "ng/mL", auclast = "day*ng/mL",
            half.life = "day"),
  tolerance_pct = 20
)
knitr::kable(cmp, align = c("l", "l", "r", "r", "r"),
             caption = paste("Typical-value simulation vs Hard 2017 Table 2",
                             "(arithmetic means).", attr(cmp, "footnote")))
```

Five rows are starred, and they are two findings rather than five: the
882 mg q6wk AUC and its Cavg (the same quantity divided by tau), and all three
half-lives. Both are discussed under *Deviations* below and neither is tuned
away. `Cmax` is reproduced in every arm, and AUC and Cavg in the other two arms.

```{r nca-gate}
pct <- function(param, reg) {
  v <- cmp$`% diff`[grepl(param, cmp$`NCA parameter`, fixed = TRUE) &
                      cmp$regimen == reg]
  stopifnot(length(v) == 1L)
  abs(as.numeric(sub("%$", "", sub("\\*$", "", v))))
}

# Deterministic simulation (zeroRe on a fixed sampling grid), so these bounds
# carry no cohort-draw noise; the headroom is against the published arithmetic
# means, which come from 19-28 subjects at 40-78% CV.
stopifnot(
  pct("Cmax", "441 q4wk") < 15,
  pct("Cmax", "882 q6wk") < 15,
  pct("Cmax", "1064 q8wk") < 15,
  pct("AUClast", "441 q4wk") < 12,
  pct("AUClast", "1064 q8wk") < 12,
  pct("Cavg", "441 q4wk") < 12,
  pct("Cavg", "1064 q8wk") < 12
)
```

## Dose linearity

The model is linear, so dose-normalised exposure over the last interval must be
the same in all three arms. The published table is not -- which is what the
882 mg q6wk deviation actually is.

```{r linearity}
sim_auc <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(regimen, sim_auc = PPORRES)
eqdose <- vapply(arms, function(a) a$eq, numeric(1))
names(eqdose) <- vapply(arms, function(a) a$regimen, character(1))

lin <- published |>
  dplyr::select(regimen, pub_auc = auclast) |>
  dplyr::left_join(sim_auc, by = "regimen") |>
  dplyr::mutate(
    `Equivalent dose (mg)` = eqdose[regimen],
    `Published AUCtau / dose` = pub_auc / `Equivalent dose (mg)`,
    `Simulated AUCtau / dose` = sim_auc / `Equivalent dose (mg)`
  ) |>
  dplyr::select(regimen, `Equivalent dose (mg)`,
                `Published AUCtau / dose`, `Simulated AUCtau / dose`)
knitr::kable(lin, digits = 2,
             caption = "Dose-normalised AUC over the interval after the last dose (day*ng/mL per mg aripiprazole equivalent).")

spread <- function(x) (max(x) - min(x)) / mean(x)
tibble::tibble(
  Source = c("Published (Table 2)", "Simulated (typical value)"),
  `Relative spread across arms` = c(spread(lin$`Published AUCtau / dose`),
                                    spread(lin$`Simulated AUCtau / dose`))
) |>
  knitr::kable(digits = 4)

# The simulated arms must agree with each other to within a couple of percent --
# that is a property of a linear model, not of the cohort.
sim_spread_pct <- 100 * spread(lin$`Simulated AUCtau / dose`)
stopifnot(sim_spread_pct < 3)
```

# Virtual cohort

```{r cohort}
# set.seed() seeds R's RNG (used for the weight draw). It does NOT seed
# rxode2's simulation RNG, whose streams are partitioned per solver thread, so
# the cohort differs between a 2-core CI runner and a 16-thread workstation.
# Every assertion below is written to hold for any cohort the model can draw.
set.seed(20260916)
n_per_arm <- 200

ev_cohort <- dplyr::bind_rows(lapply(seq_along(arms), function(i) {
  ids <- (i - 1L) * 1000L + seq_len(n_per_arm)
  # Table 1: mean +/- SD weight 89.8 +/- 16.9 kg; clipped to the earlier
  # PopPK data set's observed 43-144 kg range.
  wt <- pmin(pmax(stats::rnorm(n_per_arm, 89.8, 16.9), 43), 144)
  make_arm(arms[[i]], ids = ids, wt = wt)
}))
stopifnot(!anyDuplicated(unique(ev_cohort[, c("id", "time", "evid")])))

sim_cohort <- suppressWarnings(
  rxode2::rxSolve(ui, ev_cohort, keep = c("regimen", "WT"), addDosing = FALSE)
) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc))

# A handful of subjects with an extreme eta draw decay into solver noise in the
# far tail and return values around -5e-10 ng/mL. That is round-off, not a
# model defect, so the assertion is on the MAGNITUDE of any negative value --
# eight orders of magnitude below the cohort median, so a genuinely negative
# prediction still goes red -- and the values are then clamped so the
# percentile bands and any log axis stay well defined.
worst_negative <- min(c(0, sim_cohort$Cc))
stopifnot(worst_negative > -1e-6)
sim_cohort$Cc <- pmax(sim_cohort$Cc, 0)
```

```{r figure-1}
sim_cohort |>
  dplyr::group_by(regimen, time) |>
  dplyr::summarise(Q05 = quantile(Cc, 0.05), Q50 = median(Cc),
                   Q95 = quantile(Cc, 0.95), .groups = "drop") |>
  ggplot(aes(time / 24, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.22) +
  geom_line() +
  geom_line(data = sim_tv, aes(time / 24, Cc), colour = "firebrick",
            linetype = 2, inherit.aes = FALSE) +
  facet_wrap(~regimen) +
  labs(x = "Study day", y = "Aripiprazole (ng/mL)",
       title = "Simulated A105 cohort, 200 subjects per arm",
       caption = paste("Replicates the shape of Figure 1 of Hard 2017 (mean and",
                       "SD by regimen). Band = 5th-95th percentile, solid =",
                       "median, dashed red = typical-value profile. Spread is",
                       "wider than the published data because the paper's",
                       "8 x 8 OMEGA block off-diagonals are unpublished."))
```

The cohort median profile should sit on the typical-value profile **without a
systematic offset**. It does not sit on it point by point: with nine
uncorrelated log-normal etas, three of which have CVs above 140%, the median
subject at week 4 is not the median subject at week 40, and the resulting
shape-and-timing scatter is tens of percent at any single sampling day. The
check below is therefore on the **median signed** deviation per arm -- a
statistic that is near zero when the cohort is centred correctly and that moves
by a factor when something structural is wrong (the classic failure being
duplicate subject IDs across arms, which silently sums doses).

```{r cohort-median-check}
med <- sim_cohort |>
  dplyr::group_by(regimen, time) |>
  dplyr::summarise(Q50 = median(Cc), .groups = "drop") |>
  dplyr::inner_join(dplyr::select(sim_tv, regimen, time, Cc), by = c("regimen", "time")) |>
  dplyr::filter(time > 30 * 24) |> # after the first input has established itself
  dplyr::mutate(pct = 100 * (Q50 - Cc) / Cc)

bias <- med |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(
    `Median signed % diff` = median(pct),
    `10th pct` = quantile(pct, 0.1),
    `90th pct` = quantile(pct, 0.9),
    .groups = "drop"
  )
knitr::kable(bias, digits = 2,
             caption = "Cohort median profile vs the typical-value profile, across sampling days after day 30. The centre is unbiased; the per-day scatter is the shape-and-timing spread the large uncorrelated etas produce.")

# Centre only -- never the per-day extremes, which are one draw on one thread
# count. Realised -8.7 / +3.1 / +0.1 percent across the three arms here; 25
# leaves room for another draw while still going red on an id-collision bug
# (those inflate an arm by 100-300%) or a mis-scaled dose.
stopifnot(all(abs(bias$`Median signed % diff`) < 25))
```

# Steady-state exposure across the approved regimens

Supplemental Table 4 reports simulated median steady-state Cmin / Cmax / Cavg
after 48 weeks of dosing for the four approved regimens plus the proposed
1064 mg q8wk regimen. At true steady state a linear model's Cavg is fixed by
mass balance at `FIM * Dose / (CL/F * tau)`, so the typical-value prediction can
be written in closed form and compared with the paper's simulated medians.

```{r steady-state}
ss <- tibble::tribble(
  ~regimen,       ~eq,  ~tau_d, ~pub_cavg,
  "441 mg q4wk",   300, 28,     119,
  "662 mg q4wk",   450, 28,     183,
  "882 mg q4wk",   600, 28,     249,
  "882 mg q6wk",   600, 42,     165,
  "1064 mg q8wk",  724, 56,     154
) |>
  dplyr::mutate(
    `Closed-form Cavg,ss (ng/mL)` = 1000 * fim * eq / (cl_typ * tau_d * 24),
    `% vs published median` = 100 * (`Closed-form Cavg,ss (ng/mL)` - pub_cavg) / pub_cavg
  ) |>
  dplyr::rename("Regimen" = regimen, "Aripiprazole equivalent (mg)" = eq,
                "tau (days)" = tau_d, "Published median Cavg,ss (ng/mL)" = pub_cavg)
knitr::kable(ss, digits = 1,
             caption = "Closed-form steady-state Cavg from the encoded CL/F and FIM vs the simulated medians of ESM Supplemental Table 4.")

# The closed form runs 5-13% above the paper's simulated medians in every
# regimen; a transcription error in CL/F, FIM or the dose mapping would move it
# by tens of percent, so 20% still goes red. See Deviations.
stopifnot(all(abs(ss$`% vs published median`) < 20))
```

# Claims the paper makes about the model

```{r claims}
claims <- tibble::tibble(
  Claim = c(
    "Duration of absorption is 43 days (ESM Suppl. Table 7 D1 = 1043 h)",
    "Lag time is 3.2 days (ESM Suppl. Table 7 ALAG = 77.5 h)",
    "Total duration of input into the systemic circulation is 46 days",
    "CL/F is reduced by 23% in CYP2D6 poor metabolizers",
    "IM bioavailability relative to oral aripiprazole is 57%"
  ),
  Achieved = c(
    sprintf("%.1f days", d1_typ / 24),
    sprintf("%.1f days", tlag_typ / 24),
    sprintf("%.1f days", (d1_typ + tlag_typ) / 24),
    sprintf("%.0f%% reduction", 100 * (1 - ui$theta[["e_cyp2d6_pm_cl"]])),
    sprintf("%.0f%%", 100 * fim)
  ),
  Pass = c(
    abs(d1_typ / 24 - 43) < 1,
    abs(tlag_typ / 24 - 3.2) < 0.2,
    abs((d1_typ + tlag_typ) / 24 - 46) < 1,
    abs(100 * (1 - ui$theta[["e_cyp2d6_pm_cl"]]) - 23) < 1,
    abs(100 * fim - 57) < 1
  )
)
knitr::kable(claims)
stopifnot(is.logical(claims$Pass), !anyNA(claims$Pass), all(claims$Pass))
```

# Deviations

Three comparisons deserve comment. Two of them fall outside the 20% tolerance
of the table above; the third sits inside it but is systematic enough to be
worth stating. All three are recorded rather than tuned away.

**1. The 882 mg q6wk arm's observed AUC is ~19% below what a linear model
predicts, and is internally inconsistent with the other two arms of the same
study.** Dose-normalised AUC over the last interval in Table 2 is 11.7, 9.2 and
10.9 day*ng/mL per mg aripiprazole equivalent for the 441 q4wk, 882 q6wk and
1064 q8wk arms. A linear model cannot produce that spread, and the 2MPopPK
model is linear; the simulated values agree with each other to within
`r sprintf("%.1f%%", sim_spread_pct)`.
The published q6wk arm has n = 19-23 with 47% CV, so its standard error is
about 10% and the gap is roughly two standard errors. The model reproduces
Cmax in that arm to +1.8%, so the discrepancy sits in the AUC of the observed
cohort rather than in the encoded parameters.

**2. The half-life implied by the model is 140-154 days against a published
54-57 days.** The paper's non-compartmental half-lives were fitted over the
140-day follow-up window, of which the first 46 days still carry input from the
last injection, and their between-subject CVs are 58-78%. The model's own
terminal eigenvalue (computed above) is a `r sprintf("%.0f", t_half_beta)`-day
beta phase, driven by Vp/F = 2122 L and Q/F = 0.423 L/h. Those two parameters
were **not estimated** by the 2MPopPK model: ESM Supplemental Table 7 footnote a
fixes them at the earlier PopPK model's values, where the deep compartment was
fitted jointly with dehydro-aripiprazole. The 2MPopPK model dropped
dehydro-aripiprazole but kept the disposition parameters, so its terminal phase
is inherited rather than re-identified from these data. Use the model for the
exposure metrics it was built for (Cmax, Cavg, AUC over a dosing interval), not
for terminal half-life.

**3. The closed-form steady-state Cavg runs 5-13% above the simulated medians of
Supplemental Table 4.** The offset is largest for the q4wk regimens and smallest
for q8wk. Mass balance fixes Cavg at steady state, so the paper's simulated
medians must reflect something other than a fully-converged steady state --
plausibly the 48-week simulation horizon against the model's very slow terminal
phase, or the seven-depot ceiling of the NONMEM implementation (a 48-week q4wk
course needs 13 injections but the published model carries only seven IM
depots, whereas the q8wk course needs only seven).

# Assumptions and deviations

* **Weight centering (70 kg) comes from the earlier paper, not this one.** The
  2MPopPK paper states only that the weight effect on Vc/F was "retained" from
  the earlier PopPK model and never prints the centering value. Hard 2017
  *J Clin Psychopharmacol* Supplementary Data Content states it explicitly
  ("centered for a weight of 70 kg") and prints `VC/F = 268*(WT/70)^1.0`. That
  paper is the direct predecessor, is cited as reference 2 of the ESM, and
  supplies the values of Vp/F, Q/F and ARI(0) that the 2MPopPK model fixed, so
  it is the authoritative source for the centering constant too.

* **Ka and FIM are encoded as estimated, not fixed, against two prose
  statements to the contrary.** ESM Sect. 4 and the main paper's Discussion both
  say the oral-related parameters "were fixed to the estimates obtained from the
  earlier PopPK model". Supplemental Table 7 disagrees: `Ka` and `FIM` carry
  %RSE and 95% CIs and are not marked with the `**` fixed flag that VP/F, Q/F,
  ARI(0), FPO and the WT exponent all carry. The earlier model's values settle
  it -- its Ka is 0.574 1/h and its FIM (Formulation 2) is 0.581, whereas the
  2MPopPK values are 0.803 and 0.571. Values that differ from the ones they were
  supposedly fixed to were re-estimated. The prose most likely describes the
  base-model step, before those parameters were freed.

* **The off-diagonals of the 8 x 8 OMEGA block are unpublished.** ESM Sect. 4
  states that IIV on eight of the structural parameters was estimated "in an
  OMEGA block", but Supplemental Table 7 prints only the diagonal variances. The
  model therefore encodes diagonal IIV only. Typical-value predictions are
  unaffected; **simulated between-subject spread is wider than the paper's**,
  most visibly for the long dosing intervals where the large, uncorrelated Q/F
  (CV 145%) and Vp/F (CV 435%) variabilities are free to move independently of
  clearance. Do not read the cohort percentile bands above as a reproduction of
  the paper's VPCs.

* **The seven IM depots are encoded as one lagged zero-order input to
  `central`.** They are mathematically equivalent by superposition (verified by
  the overlapping-input mass-balance check above), and the encoded form has no
  seven-injection ceiling. The trade-off is that `f(central)` now carries FIM,
  so a hypothetical intravenous dose written to `central` would also be scaled
  by 0.571; the published data set contains no intravenous route.

* **`ARI(0)` is encoded as an initial condition on `central`, not as an additive
  offset on the prediction.** The `(0)` notation is NONMEM's initial-amount
  syntax and the earlier paper's abbreviation list reads "ARI(0), baseline
  amount of ARI"; both tables report it in ng/mL, so the encoding is
  `central(0) <- c0 * vc / 1000`. It decays with the disposition, which is the
  behaviour a residual pre-study oral dose should have. Its IIV variance of 5.89
  is very large (the table's own CV is 1898%); users simulating the first weeks
  after a first injection should be aware that a small number of virtual
  subjects will start with a substantial residual concentration.

* **`FPO`'s IIV was fixed to zero** (ESM Sect. 4), so no eta is attached to
  `lfdepot`.

* **Inter-occasion variability is encoded over seven occasions.** The paper
  reports a single IOV variance on `D1` and its model carried seven IM depots,
  so seven occasion slots are encoded with occasions 2-7 fixed to the
  occasion-1 estimate, the NONMEM `OMEGA BLOCK(1) SAME` idiom. A data set with
  more than seven injections should either recycle `OCC` or accept `OCC = 0`
  (no IOV) on the later records.

* **No covariate effects beyond weight-on-Vc/F and CYP2D6-on-CL/F.** The full
  model also carried surface area on D1, ALAG and FIM, and race and sex on
  CL/F; ESM Supplemental Table 3 reports all seven. None of the five additional
  effects reduced the objective function value and all were dropped, so they are
  not encoded. They are recorded in the model file's `covariateData` notes only
  where they overlap a retained covariate.

* **The virtual cohort's weight distribution** is drawn as
  `N(89.8, 16.9)` kg (Table 1 of the phase I study) and clipped to the
  43-144 kg range the earlier PopPK data set spanned. The simulated cohorts are
  all CYP2D6 non-poor-metabolizers and all flagged `STUDY_A105 = 1`, matching
  the arm being replicated.
