---
title: "Cefotaxime (Hartman 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cefotaxime (Hartman 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Hartman_2022_cefotaxime"))$reference`
* Description: `r rxode2::rxode(readModelDb("Hartman_2022_cefotaxime"))$description`
* Article: <https://doi.org/10.1093/jac/dkac095> (open access, PMC9155601)
* Supplement: Supplementary data (methods, Figures S1-S6 and the final NONMEM
  control stream), available from the article page at JAC Online and from
  Europe PMC.

Hartman et al. characterised the pharmacokinetics of intravenous cefotaxime in
critically ill children and used the final model to evaluate current and
alternative dosing regimens against a 100% fT>MIC target (trough unbound
concentration above the MIC) for MICs of 0.5, 2 and 4 mg/L.

# Population

The analysis included 52 children (median postnatal age 1.61 years, range
0.03-17.69 years; median weight 10.95 kg, range 2.7-80 kg; 38.5% female)
admitted to the level-3 paediatric intensive care unit of the Radboudumc
(Nijmegen, the Netherlands) between June 2017 and May 2019 as part of the
POPSICLE study (NCT03248349). The main reasons for admission were respiratory
failure (63.5%), neurological impairment (17.3%) and circulatory failure
(9.6%); 90.4% were mechanically ventilated and 36.5% received vasopressive
co-medication during the study. Most children received the prophylactic dose
of 100 mg/kg/day (maximum 4 g/day, selective decontamination of the digestive
tract); 11 (21.2%) received the therapeutic dose of 150 mg/kg/day (maximum 12
g/day), as three or four intravenous push doses per day. In total 479 total
plasma cefotaxime concentrations (median 10 per patient) entered the final
model. Demographics are from Table 1 of Hartman 2022; the cohort was
predominantly Caucasian (>90%).

The same information is available programmatically:

```{r population}
str(rxode2::rxode(readModelDb("Hartman_2022_cefotaxime"))$population)
```

# Source trace

Every `ini()` value carries an in-file comment pointing to its source. The
final NONMEM control stream in the supplement confirms the structure
(ADVAN5, two compartments, weight on CL and V1 only, `Y = F*EXP(EPS(1))`).

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL at 10.95 kg) | log(2.8) L/h | Table 2, CLpop |
| `lvc` (V1 at 10.95 kg) | log(2.62) L | Table 2, V1pop |
| `lvp` (V2) | log(1.55) L | Table 2, V2pop |
| `lq` (Q) | log(1.15) L/h | Table 2, Qpop |
| `e_wt_cl` | 1.11 | Table 2, Theta1 |
| `e_wt_vc` | 1.18 | Table 2, Theta2 |
| `etalcl` variance | 0.359 | Table 2, IIV CL; 65.7% CV in Results |
| `etalcl`-`etalvc` covariance | 0.305 | Table 2, 'Block matrix' |
| `etalvc` variance | 0.581 | Table 2, IIV V1; 88.8% CV in Results |
| `expSd` | sqrt(0.307) | Table 2, 'Proportional error' (raw $SIGMA variance) |
| `CL = CLpop * (WT/10.95)^Theta1 * exp(eta)` | n/a | Table 2 equation; control stream $PK |
| `V1 = V1pop * (WT/10.95)^Theta2 * exp(eta)` | n/a | Table 2 equation; control stream $PK |
| `V2 = V2pop`, `Q = Qpop` (no weight scaling) | n/a | Control stream $PK (`V2 = TVV2`, `Q = TVQ`) |
| `d/dt(central)`, `d/dt(peripheral1)` | n/a | Control stream $DES (ADVAN5 K10, K12, K21) |
| `Cc ~ lnorm(expSd)` | n/a | Control stream $ERROR `Y = F*EXP(EPS(1))`; supplementary results |

## Checks against the published numbers

The Discussion quotes the typical clearance and total volume per kilogram at
the cohort median weight (0.26 L/kg/h and 0.38 L/kg), and the Results quote the
final-model IIV as 65.7% (CL) and 88.8% (V1). These are deterministic
functions of the `ini()` values and are checked exactly.

```{r published-checks}
ui <- rxode2::rxode(readModelDb("Hartman_2022_cefotaxime"))
th <- ui$theta
om <- ui$omega

cl_per_kg <- exp(th[["lcl"]]) / 10.95
vd_per_kg <- (exp(th[["lvc"]]) + exp(th[["lvp"]])) / 10.95
cv_cl <- 100 * sqrt(exp(om["etalcl", "etalcl"]) - 1)
cv_vc <- 100 * sqrt(exp(om["etalvc", "etalvc"]) - 1)
corr_cl_vc <- om["etalcl", "etalvc"] / sqrt(om["etalcl", "etalcl"] * om["etalvc", "etalvc"])

knitr::kable(
  data.frame(
    Quantity = c("CL per kg (L/kg/h)", "Vd per kg (L/kg)", "IIV CL (%CV)", "IIV V1 (%CV)", "CL-V1 correlation"),
    Model = signif(c(cl_per_kg, vd_per_kg, cv_cl, cv_vc, corr_cl_vc), 3),
    Paper = c("0.26", "0.38", "65.7", "88.8", "not reported")
  ),
  caption = "Typical values and IIV implied by the model against the values quoted in Hartman 2022."
)

stopifnot(
  abs(round(cl_per_kg, 2) - 0.26) < 1e-9,
  abs(round(vd_per_kg, 2) - 0.38) < 1e-9,
  abs(round(cv_cl, 1) - 65.7) < 1e-9,
  abs(round(cv_vc, 1) - 88.8) < 1e-9
)
```

# Dose evaluation in typical patients (Figure 2 and Figure S6)

The paper's dose evaluations use five typical patients with a normal weight
for age (1 week 3.775 kg, 1 year 9.75 kg, 4 years 17.125 kg, 8 years 26.75 kg,
18 years 60.5 kg), a maximum daily dose of 12 g, unbound concentrations taken
as 60% of total, and the trough 24 h after the first dose as the surrogate for
fT>MIC. The continuous-infusion regimens start with a 25 mg/kg loading dose
(maximum 1000 mg).

```{r typical-events}
typical <- tibble::tibble(
  patient = c("1 week - 3.775 kg", "1 year - 9.75 kg", "4 year - 17.125 kg", "8 year - 26.75 kg", "18 years - 60.5 kg"),
  WT = c(3.775, 9.75, 17.125, 26.75, 60.5)
)

regimens <- tibble::tribble(
  ~regimen, ~mgkgday, ~tau, ~dur,
  "150 mg/kg/day bolus q8h", 150, 8, 0,
  "200 mg/kg/day bolus q6h", 200, 6, 0,
  "150 mg/kg/day bolus q4h", 150, 4, 0,
  "200 mg/kg/day bolus q4h", 200, 4, 0,
  "150 mg/kg/day 60-min infusion q8h", 150, 8, 1,
  "200 mg/kg/day 60-min infusion q6h", 200, 6, 1,
  "150 mg/kg/day 120-min infusion q8h", 150, 8, 2,
  "200 mg/kg/day 120-min infusion q6h", 200, 6, 2,
  "150 mg/kg/day 240-min infusion q8h", 150, 8, 4,
  "200 mg/kg/day 240-min infusion q6h", 200, 6, 4,
  "100 mg/kg/day continuous + 25 mg/kg LD", 100, 24, 24,
  "150 mg/kg/day continuous + 25 mg/kg LD", 150, 24, 24,
  "200 mg/kg/day continuous + 25 mg/kg LD", 200, 24, 24
)

obs_times <- sort(unique(c(seq(0, 24, by = 0.1), 24)))

make_typical_events <- function(regimen_row, typical, obs_times) {
  rows <- vector("list", nrow(typical))
  for (i in seq_len(nrow(typical))) {
    wt <- typical$WT[i]
    daily <- min(regimen_row$mgkgday * wt, 12000)
    dose_times <- seq(0, 24 - regimen_row$tau, by = regimen_row$tau)
    amt <- daily / length(dose_times)
    doses <- tibble::tibble(time = dose_times, amt = amt, evid = 1L, rate = 0)
    if (regimen_row$dur > 0) {
      doses$rate <- amt / regimen_row$dur
    }
    if (regimen_row$dur == 24) {
      # Continuous infusion: 25 mg/kg (maximum 1000 mg) loading bolus plus the
      # daily dose infused over 24 h.
      doses <- dplyr::bind_rows(
        tibble::tibble(time = 0, amt = min(25 * wt, 1000), evid = 1L, rate = 0),
        doses
      )
    }
    obs <- tibble::tibble(time = obs_times, amt = NA_real_, evid = 0L, rate = NA_real_)
    rows[[i]] <- dplyr::bind_rows(doses, obs) |>
      dplyr::mutate(patient = typical$patient[i], WT = wt, cmt = "central")
  }
  dplyr::bind_rows(rows) |> dplyr::mutate(regimen = regimen_row$regimen)
}

typical_events <- dplyr::bind_rows(lapply(seq_len(nrow(regimens)), function(j) {
  make_typical_events(regimens[j, ], typical, obs_times)
})) |>
  dplyr::mutate(id = as.integer(factor(paste(regimen, patient), levels = unique(paste(regimen, patient))))) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(length(unique(typical_events$id)) == nrow(regimens) * nrow(typical))
```

```{r typical-solve}
mod <- readModelDb("Hartman_2022_cefotaxime")
mod_typical <- rxode2::zeroRe(mod)
# zeroRe() leaves a multi-subject solve with no omega; rxode2 warns about
# exactly that, which is the intent here.
sim_typical <- suppressWarnings(rxode2::rxSolve(
  mod_typical,
  events = typical_events,
  keep = c("regimen", "patient")
)) |>
  as.data.frame() |>
  dplyr::mutate(Cu = 0.6 * Cc)

# Pre-dose rows are exactly zero and cannot be drawn on a log axis.
sim_plot <- sim_typical |>
  dplyr::mutate(Cu = dplyr::if_else(Cu > 0, Cu, NA_real_))
```

```{r figure-2, fig.height = 4}
mic_lines <- data.frame(MIC = c(0.5, 2, 4), label = c("MIC 0.5 mg/L", "MIC 2 mg/L", "MIC 4 mg/L"))
fig2_regimens <- c(
  "200 mg/kg/day bolus q6h",
  "200 mg/kg/day bolus q4h",
  "200 mg/kg/day continuous + 25 mg/kg LD"
)
sim_plot |>
  dplyr::filter(regimen %in% fig2_regimens) |>
  dplyr::mutate(
    regimen = factor(regimen, levels = fig2_regimens),
    patient = factor(patient, levels = typical$patient)
  ) |>
  ggplot(aes(time, Cu, colour = patient)) +
  geom_line(na.rm = TRUE) +
  geom_hline(data = mic_lines, aes(yintercept = MIC, linetype = label)) +
  facet_wrap(~regimen) +
  scale_y_log10(limits = c(0.01, 1000)) +
  labs(
    x = "Time (h)", y = "Unbound cefotaxime (mg/L)", colour = "Typical patient", linetype = NULL,
    title = "Typical-patient unbound concentrations, three regimens",
    caption = "Replicates Figure 2 of Hartman 2022 (unbound = 0.6 x total)."
  ) +
  theme(legend.position = "bottom")
```

```{r figure-s6, fig.height = 10}
sim_plot |>
  dplyr::mutate(
    regimen = factor(regimen, levels = regimens$regimen),
    patient = factor(patient, levels = typical$patient)
  ) |>
  ggplot(aes(time, Cu, colour = patient)) +
  geom_line(na.rm = TRUE) +
  geom_hline(data = mic_lines, aes(yintercept = MIC, linetype = label)) +
  facet_wrap(~regimen, ncol = 2) +
  scale_y_log10(limits = c(0.01, 1000)) +
  labs(
    x = "Time (h)", y = "Unbound cefotaxime (mg/L)", colour = "Typical patient", linetype = NULL,
    title = "All evaluated regimens",
    caption = "Replicates Figure S6 of Hartman 2022 (unbound = 0.6 x total)."
  ) +
  theme(legend.position = "bottom")
```

## Trough concentrations at 24 h

The paper reports its conclusions graphically. The reference values below were
digitised by the maintainers from the 24 h trough points of the bolus panels
of Figure S6 (log axis; reading precision roughly +/-10%) and from the
plateaus of the continuous-infusion panels.

```{r trough-table}
troughs <- sim_typical |>
  dplyr::filter(time == 24) |>
  dplyr::select(regimen, patient, Cu)

digitised <- tibble::tribble(
  ~regimen, ~patient, ~Cu_fig,
  "150 mg/kg/day bolus q8h", "1 week - 3.775 kg", 3.0,
  "150 mg/kg/day bolus q8h", "1 year - 9.75 kg", 1.0,
  "150 mg/kg/day bolus q8h", "4 year - 17.125 kg", 0.5,
  "150 mg/kg/day bolus q8h", "8 year - 26.75 kg", 0.3,
  "150 mg/kg/day bolus q8h", "18 years - 60.5 kg", 0.15,
  "200 mg/kg/day bolus q6h", "1 week - 3.775 kg", 6.0,
  "200 mg/kg/day bolus q6h", "1 year - 9.75 kg", 2.5,
  "200 mg/kg/day bolus q6h", "4 year - 17.125 kg", 1.4,
  "200 mg/kg/day bolus q6h", "8 year - 26.75 kg", 1.0,
  "200 mg/kg/day bolus q6h", "18 years - 60.5 kg", 0.6,
  "150 mg/kg/day bolus q4h", "1 week - 3.775 kg", 6.0,
  "150 mg/kg/day bolus q4h", "1 year - 9.75 kg", 3.0,
  "150 mg/kg/day bolus q4h", "4 year - 17.125 kg", 2.0,
  "150 mg/kg/day bolus q4h", "8 year - 26.75 kg", 1.6,
  "150 mg/kg/day bolus q4h", "18 years - 60.5 kg", 1.35,
  "200 mg/kg/day bolus q4h", "1 week - 3.775 kg", 8.5,
  "200 mg/kg/day bolus q4h", "1 year - 9.75 kg", 3.8,
  "200 mg/kg/day bolus q4h", "4 year - 17.125 kg", 2.5,
  "200 mg/kg/day bolus q4h", "8 year - 26.75 kg", 2.1,
  "200 mg/kg/day bolus q4h", "18 years - 60.5 kg", 1.8,
  "200 mg/kg/day continuous + 25 mg/kg LD", "1 week - 3.775 kg", 22,
  "200 mg/kg/day continuous + 25 mg/kg LD", "18 years - 60.5 kg", 17
)

trough_cmp <- digitised |>
  dplyr::left_join(troughs, by = c("regimen", "patient")) |>
  dplyr::mutate(pct_diff = 100 * (Cu / Cu_fig - 1))
stopifnot(nrow(trough_cmp) == nrow(digitised), !anyNA(trough_cmp$Cu))

trough_cmp |>
  dplyr::mutate(Cu = signif(Cu, 3), pct_diff = round(pct_diff, 1)) |>
  dplyr::rename(
    "Regimen" = regimen,
    "Typical patient" = patient,
    "Figure S6 (mg/L)" = Cu_fig,
    "Model (mg/L)" = Cu,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(caption = "Unbound trough at 24 h: model versus values digitised from Figure S6.")
```

All typical-patient troughs fall within the digitising precision of the figure.
The comparison is deterministic (typical values, no random effects), so the
bound below does not depend on the simulation seed; a transcription error in a
clearance, volume or exponent moves these troughs by far more than 25%.

```{r trough-gate}
stopifnot(max(abs(trough_cmp$pct_diff)) < 25)
```

## Dosing conclusions (Results and Table 3)

```{r claims}
cu24 <- function(reg, pat_age) {
  v <- troughs$Cu[troughs$regimen == reg & startsWith(troughs$patient, pat_age)]
  if (length(v) != 1L) stop("no unique trough for '", reg, "' / '", pat_age, "'")
  v
}
ages <- c("1 week", "1 year", "4 year", "8 year", "18 years")
q8_150 <- vapply(ages, cu24, numeric(1), reg = "150 mg/kg/day bolus q8h")
q6_200 <- vapply(ages, cu24, numeric(1), reg = "200 mg/kg/day bolus q6h")
q4_200 <- vapply(ages, cu24, numeric(1), reg = "200 mg/kg/day bolus q4h")
ci_200 <- vapply(ages, cu24, numeric(1), reg = "200 mg/kg/day continuous + 25 mg/kg LD")

claims <- tibble::tribble(
  ~claim, ~holds, ~deviation,
  "150 mg/kg/day: trough < 0.5 mg/L in patients > 4 years (8 and 18 years)", all(q8_150[c("8 year", "18 years")] < 0.5), FALSE,
  "200 mg/kg/day q6h: trough > 0.5 mg/L in all typical patients", all(q6_200 > 0.5), FALSE,
  "200 mg/kg/day q6h: trough > 2 mg/L in patients < 1 year (1 week)", q6_200[["1 week"]] > 2, FALSE,
  "200 mg/kg/day q4h: trough > 2 mg/L in all typical patients", all(q4_200 > 2), TRUE,
  "200 mg/kg/day q4h: trough > 2 mg/L in patients 1 week to 8 years", all(q4_200[c("1 week", "1 year", "4 year", "8 year")] > 2), FALSE,
  "200 mg/kg/day q4h: trough > 4 mg/L in patients < 1 year (1 week)", q4_200[["1 week"]] > 4, FALSE,
  "200 mg/kg/day q4h: trough < 4 mg/L in patients > 1 year (CI needed for MIC 4)", all(q4_200[c("4 year", "8 year", "18 years")] < 4), FALSE,
  "25 mg/kg LD + 200 mg/kg/day continuous: > 4 mg/L in all typical patients", all(ci_200 > 4), FALSE
)
claims |>
  dplyr::rename("Claim" = claim, "Reproduced" = holds, "Known deviation" = deviation) |>
  knitr::kable(caption = "Dosing conclusions of Hartman 2022 evaluated on the typical-patient simulations.")
stopifnot(all(claims$holds[!claims$deviation]))
# The one known deviation sits at the MIC line, not far from it: the 18-year-old
# trough on 200 mg/kg/day q4h is just under 2 mg/L, as it is in Figure S6.
stopifnot(q4_200[["18 years"]] > 1.6, q4_200[["18 years"]] < 2)
```

One conclusion is not reproduced exactly. The Results state that 200 mg/kg/day
every 4 h gives adequate troughs for MICs up to 2 mg/L in all typical
patients, but the 18-year-old (60.5 kg) trough is
`r signif(q4_200[["18 years"]], 3)` mg/L unbound. Figure S6 of the paper shows
the same thing: the 18-year-old points of the 200 mg/kg/day q4h panel sit just
under the 2 mg/L reference line (digitised at about 1.8 mg/L, table above). The
statement in the text rounds a near-miss up; the model agrees with the
published figure.

# Stochastic simulation of the study population

A virtual cohort of 200 children per arm is drawn with a log-normal weight
distribution matched to the Table 1 median (10.95 kg) and interquartile range
(5.2-28.5 kg), truncated to the observed 2.7-80 kg range. Two arms mirror the
study doses: the prophylactic 100 mg/kg/day (maximum 4 g/day) and the
therapeutic 150 mg/kg/day (maximum 12 g/day), both as intravenous push doses
every 6 h for 3 days.

```{r cohort}
rxode2::rxSetSeed(2022)
set.seed(2022)
n_per_arm <- 200

draw_wt <- function(n) {
  mu <- log(10.95)
  sdlog <- (log(28.5) - log(5.2)) / (2 * qnorm(0.75))
  out <- numeric(0)
  while (length(out) < n) {
    w <- stats::rlnorm(n, mu, sdlog)
    out <- c(out, w[w >= 2.7 & w <= 80])
  }
  out[seq_len(n)]
}

make_cohort <- function(n, mgkgday, maxday, label, id_offset) {
  subj <- tibble::tibble(id = id_offset + seq_len(n), WT = draw_wt(n), treatment = label)
  dose_times <- seq(0, 66, by = 6)
  obs_times <- sort(unique(c(seq(0, 72, by = 0.5), seq(66, 72, by = 0.05))))
  doses <- tidyr::crossing(subj, time = dose_times) |>
    dplyr::mutate(amt = pmin(mgkgday * WT, maxday) / 4, evid = 1L)
  obs <- tidyr::crossing(subj, time = obs_times) |>
    dplyr::mutate(amt = NA_real_, evid = 0L)
  dplyr::bind_rows(doses, obs) |>
    dplyr::mutate(cmt = "central") |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  make_cohort(n_per_arm, 100, 4000, "100 mg/kg/day q6h", 0L),
  make_cohort(n_per_arm, 150, 12000, "150 mg/kg/day q6h", n_per_arm)
)
stopifnot(length(unique(events$id)) == 2 * n_per_arm)
```

```{r simulate}
sim <- rxode2::rxSolve(mod, events = events, keep = c("treatment", "WT")) |>
  as.data.frame()
```

```{r vpc, fig.height = 4}
sim |>
  dplyr::filter(time <= 24) |>
  dplyr::group_by(treatment, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~treatment) +
  scale_y_log10() +
  labs(
    x = "Time after first dose (h)", y = "Total cefotaxime (mg/L)",
    title = "Simulated total plasma cefotaxime, first 24 h",
    caption = "Median and 90% prediction interval of the individual predictions."
  )
```

The paper reports an over 10-fold range of observed trough concentrations and
individual clearances between 0.02 and 0.66 L/kg/h (Figure S2). The simulated
cohort reproduces that breadth; the check is on robust quantiles, not the
extremes, so it holds for any draw.

```{r spread}
trough_sim <- sim |> dplyr::filter(time == 24)
cl_kg <- sim |>
  dplyr::distinct(id, .keep_all = TRUE) |>
  dplyr::mutate(cl_kg = cl / WT)
spread <- c(
  trough_ratio_p95_p05 = unname(quantile(trough_sim$Cc, 0.95) / quantile(trough_sim$Cc, 0.05)),
  cl_kg_median = median(cl_kg$cl_kg),
  cl_kg_p05 = unname(quantile(cl_kg$cl_kg, 0.05)),
  cl_kg_p95 = unname(quantile(cl_kg$cl_kg, 0.95))
)
signif(spread, 3)
stopifnot(
  spread[["trough_ratio_p95_p05"]] > 10,
  spread[["cl_kg_median"]] > 0.15, spread[["cl_kg_median"]] < 0.4,
  spread[["cl_kg_p05"]] > 0.02, spread[["cl_kg_p95"]] < 1.0
)
```

# PKNCA validation

The paper does not report NCA parameters, so the NCA check is internal: at
steady state (the last 6-h dosing interval of day 3) the AUC over the dosing
interval must equal dose / CL for each simulated subject.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc), time >= 66) |>
  dplyr::select(id, time, Cc, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)
dose_df <- events |>
  dplyr::filter(evid == 1, time == 66) |>
  dplyr::select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id, route = "intravascular")

intervals <- data.frame(start = 66, end = 72, cmax = TRUE, cmin = TRUE, auclast = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_res$result) |>
  dplyr::select(treatment, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

closed_form <- sim |>
  dplyr::distinct(id, .keep_all = TRUE) |>
  dplyr::select(id, cl) |>
  dplyr::left_join(dose_df |> dplyr::select(id, amt), by = "id") |>
  dplyr::mutate(auc_tau = amt / cl)

chk <- nca_wide |>
  dplyr::left_join(closed_form, by = "id") |>
  dplyr::mutate(pct_diff = 100 * (auclast / auc_tau - 1))

chk |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    "Median Cmax,ss (mg/L)" = signif(median(cmax), 3),
    "Median Cmin,ss (mg/L)" = signif(median(cmin), 3),
    "Median AUCtau (mg*h/L)" = signif(median(auclast), 3),
    "Median dose/CL (mg*h/L)" = signif(median(auc_tau), 3),
    "Median % difference" = round(median(pct_diff), 2),
    .groups = "drop"
  ) |>
  dplyr::rename("Treatment" = treatment) |>
  knitr::kable(caption = "PKNCA steady-state interval (66-72 h) versus the dose/CL identity.")

signif(quantile(abs(chk$pct_diff), c(0.5, 0.9, 1)), 3)
# The two sides share each subject's own parameters, so the difference is
# numerical only (trapezoidal error over the 0.05-h grid on the fast
# distribution phase after a bolus); a wrong dose, unit or clearance would
# move it by tens of percent.
stopifnot(
  abs(median(chk$pct_diff)) < 0.5,
  quantile(abs(chk$pct_diff), 0.9) < 2
)
```

# Assumptions and deviations

- **Residual error scale and form.** Table 2 lists 'Proportional error 0.307'
  without a scale. The control stream codes `Y = F*EXP(EPS(1))` with
  `$SIGMA 0.311`, and the supplement states that the error model was
  implemented log-normally so that predictions could not fall below zero. The
  maintainers read 0.307 as the raw `$SIGMA` variance (the IIV rows of the same
  table are raw omega-squared values, and the `$SIGMA` initial estimate 0.311
  sits beside it) and encode `Cc ~ lnorm(expSd)` with `expSd = sqrt(0.307) =
  0.554`. Under FOCE-I on untransformed data this is equivalent to a
  proportional error of the same SD during estimation; `lnorm()` follows the
  authors' stated intent and keeps simulated observations positive.
- **Bootstrap bias column.** The '% bias' entries of Table 2 cannot all be
  reproduced from the printed estimate and bootstrap median (for example
  Theta2: estimate 1.18, bootstrap median 1.11, printed bias -0.9%). The final
  model estimates are used throughout. The main text gives 491/500 successful
  bootstrap runs and the supplement 481/500; neither affects the model.
- **Figure 2, right panel.** The panel is titled '200 mg/kg/day, continuous
  infusion with a 25 mg/kg loading dose', but its plateau (about 10 mg/L
  unbound) matches the 100 mg/kg/day continuous-infusion panel of Figure S6.
  The 200 mg/kg/day panel of Figure S6 plateaus at about 17-22 mg/L, which the
  model reproduces (table above). The replication of Figure 2 here uses 200
  mg/kg/day as titled.
- **Weight.** `WT` is the weight at the start of the ICU admission (control
  stream `$INPUT`), so it is a per-patient constant. Peripheral volume and
  inter-compartmental clearance are not weight-scaled in the published model;
  simulations for adolescents and neonates therefore keep V2 and Q at the
  10.95 kg typical values, as the authors did.
- **Unbound fraction.** The model predicts total plasma cefotaxime. The 60%
  unbound fraction the authors assumed for the target-attainment evaluation is
  applied in this vignette only (`Cu = 0.6 * Cc`), not in the model.
- **Typical-patient dosing details.** The 12 g/day maximum and the 1000 mg
  maximum loading dose are applied as stated in the Methods. The doses in
  Figure S6 appear to start shortly after time zero; here the first dose is at
  time zero and the trough is read 24 h after the first dose, as described in
  the Methods.
- **Virtual cohort.** The weight distribution is a truncated log-normal fitted
  to the Table 1 median and IQR; age, which is not a model covariate, is not
  simulated.
- **Control stream comments.** The `$INPUT` comments of the published control
  stream mention ceftriaxone and PERFORM patients; these are leftovers from the
  authors' companion ceftriaxone analysis and do not affect the model. No
  correction notice for the article was found as of 2026-10-01.
