---
title: "YPEG-rhGH (He 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{YPEG-rhGH (He 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("He_2025_YPEGrhGH"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.3389/fphar.2025.1651323>
* Supplementary material (final-model equations, bootstrap tables, study
  designs):
  <https://www.frontiersin.org/articles/10.3389/fphar.2025.1651323/full#supplementary-material>

YPEG-rhGH is a Y-shape branched 40 kDa PEGylated recombinant human growth
hormone (Xiamen Amoytop Biotech). He 2025 fitted a two-compartment model with
first-order subcutaneous absorption and Michaelis-Menten elimination to the
pooled pharmacokinetic data of 16 healthy elderly subjects and 36 healthy young
adult men, then fitted an indirect-response (IDR) model in which YPEG-rhGH
stimulates production of insulin-like growth factor 1 (IGF-1), expressed as the
ratio to each subject's own baseline.

# Population

The analysis pooled two Chinese phase-1 studies (source Table 1, Table 2 and
Supplementary Table 3):

* **TB2208GH (elderly)** -- single-arm, open-label, single-centre. 16 healthy
  subjects aged 65-74 years (median 66.0, 75% female, median weight 59.0 kg,
  BMI 18-30 kg/m^2, baseline IGF-1 standard-deviation score < 0). YPEG-rhGH
  30 ug/kg subcutaneously every 2 weeks for 23 weeks (12 injections), followed
  by 5 weeks of observation.
* **TB1010GH (healthy adults)** -- phase-1, multicentre, randomised. 36 healthy
  men aged 21-44 years (median 29.0, median weight 66.4 kg). Single
  subcutaneous doses of 10, 30, 60, 120 or 200 ug/kg after a daily-rhGH
  comparator phase and a 14-day washout.

813 of 884 pharmacokinetic samples (71 below the quantification limit were
excluded) and all 980 IGF-1 measurements entered the analysis. Sex was screened
but not retained: the two studies are essentially confounded with sex (all 36
adults were male, 12 of 16 elderly subjects were female), so the authors kept
the more direct AGE and WEIGHT covariates instead (Discussion, limitations).

The same information is available programmatically from the model's
`population` metadata:

```{r population}
str(ui$population)
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/He_2025_YPEGrhGH.R` carries an in-file comment
naming its source location. They are collected here for review. The **point
estimates** are Table 3 (PK) and Table 4 (PD); the **covariate centring values
and the final PD differential equation** are only in the Supplementary
Material, under the headings "PopPK model equations" and "PopPK/PD model
equations".

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` | `log(0.01086)` 1/h | Table 3, `Ka` (RSE 8.49%) |
| `lvc` | `log(2.387)` L | Table 3, `V1/F` (RSE 15.54%) |
| `lvp` | `log(21.42)` L | Table 3, `V2/F` (RSE 20.48%) |
| `lq` | `log(0.05575)` L/h | Table 3, `CL2/F` (RSE 10.15%) |
| `lvmax` | `log(80.13)` ug/h | Table 3, `Vmax` (RSE 7.82%) |
| `lkm` | `log(70.08)` ug/L | Table 3, `Km` (RSE 13.78%) |
| `e_age_ka` | `-0.6957` | Table 3, `AGE on Ka` (RSE 28.49%) |
| `e_age_vc` | `1.470` | Table 3, `AGE on V1/F` (RSE 24.69%) |
| `e_age_vmax` | `0.7989` | Table 3, `AGE on Vmax` (RSE 18.37%) |
| `e_wt_vmax` | `2.202` | Table 3, `WEIGHT on Vmax` (RSE 18.05%) |
| AGE centred at 35.5 years | n/a | Supplement, PopPK model equations: `Ka = 0.01086 x (AGE/35.5)^-0.6957` |
| WEIGHT centred at 70 kg | n/a | Supplement, PopPK model equations: `Vmax = 80.13 x (AGE/35.5)^0.7989 x (WEIGHT/70)^2.202` |
| Covariate functional form (power model) | n/a | Equation 1: `theta_i = theta_TV x (Cov_i / Cov_median)^theta_x` |
| `lkin` | `log(0.023)` /h | Table 4, `Kin` (RSE 14.79%) |
| `lkout` | `log(0.023)` 1/h | Table 4, `Kout` (RSE 15.50%) |
| `lemax` | `log(2.245)` | Table 4, `Emax` (RSE 21.80%) |
| `lec50` | `log(50.74)` ng/mL | Table 4, `EC50` (RSE 37.68%) |
| `hill` | `fixed(1)` | Results 3.3.1: "when gamma was fixed at 1" |
| `d/dt(igf1)` | n/a | Supplement, PopPK/PD model equations: `dR/dt = 0.023 x (1 + 2.245 x C/(50.74 + C)) - 0.023 x R` |
| Structure (2-cmt, 1st-order SC absorption, MM elimination, IDR on production) | n/a | Figure 1 (model framework diagram); Methods 2.4.1 |
| `etalka` | `fixed(0.3)` | Supplementary Table 1, `omega(Ka), %` = 54.77 with no RSE / no CI in **both** the final-model and bootstrap columns |
| `etalvc` | `0.94245` | Table 3, IIV `V1/F` = 97.08% (RSE 24.30%) |
| `etalkm` | `fixed(0.3)` | Supplementary Table 1, `omega(Km), %` = 54.77 with no RSE / no CI in **both** columns |
| `etalkout` | `0.010547` | Table 4, IIV `Kout` = 10.27% (RSE 3.901%) |
| `etalec50` | `0.620314` | Table 4, IIV `EC50` = 78.76% (RSE 30.66%) |
| `propSd` | `0.2894` | Table 3, residual `sigma (prop)` = 28.94% (RSE 2.88%) |
| `addSd_IGF1ratio` | `0.1320` | Table 4, residual `sigma (add)` = 13.20 (RSE 6.13%) |

# Virtual cohort

Original observed data are not publicly available. The cohorts below match the
covariate distributions of source Table 2 (means and SDs, truncated at the
reported minima and maxima).

```{r cohort}
# set.seed() seeds R's RNG, which drives the covariate draws below. It does NOT
# seed rxode2's simulation RNG, and rxode2's streams are partitioned per solver
# thread -- a 2-thread CI runner draws a different eta cohort than a 16-thread
# workstation. Every assertion downstream is therefore written to hold for any
# cohort the model can produce.
set.seed(20251125)

rtrunc_norm <- function(n, mean, sd, lo, hi) {
  pmin(pmax(stats::rnorm(n, mean, sd), lo), hi)
}

n_elderly <- 100L
n_adult   <- 60L

# Elderly (TB2208GH), Table 2: AGE 67.44 (3.01) [65.00, 74.00];
# WT 60.54 (7.31) [51.10, 76.10].
elderly_subj <- tibble(
  id  = seq_len(n_elderly),
  AGE = rtrunc_norm(n_elderly, 67.44, 3.01, 65.00, 74.00),
  WT  = rtrunc_norm(n_elderly, 60.54, 7.31, 51.10, 76.10)
)

# Healthy adults (TB1010GH), Table 2: AGE 30.78 (6.62) [21.00, 44.00];
# WT 65.08 (6.16) [52.40, 79.60].
make_adult_subj <- function(n, id_offset) {
  tibble(
    id  = id_offset + seq_len(n),
    AGE = rtrunc_norm(n, 30.78, 6.62, 21.00, 44.00),
    WT  = rtrunc_norm(n, 65.08, 6.16, 52.40, 79.60)
  )
}
```

The event tables are built explicitly (one row per dose, one row per
observation). Observation rows point at the **ODE state** `central` and carry
`dvid = 1` so that rxode2 maps them onto the `Cc` endpoint; `dvid = 2` rows map
onto the `IGF1ratio` endpoint. Both algebraic observables are returned as
columns on every output row, so the PD time course can be read off the `dvid = 1`
rows too.

```{r events}
# The elderly trial's actual PK sampling grid, relative to a dose
# (Supplementary Table 3).
sparse_grid <- c(0, 8, 24, 72, 96, 120, 168, 336)

make_events <- function(subj, dose_amt, ii, addl, obs_times, dvid = 1L,
                        cmt_obs = "central") {
  doses <- subj |>
    tidyr::crossing(dose_index = seq_len(addl + 1L)) |>
    mutate(
      time = (dose_index - 1L) * ii,
      amt  = dose_amt(WT),
      evid = 1L,
      cmt  = "depot",
      dvid = NA_integer_
    ) |>
    select(-dose_index)
  obs <- subj |>
    tidyr::crossing(time = obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = cmt_obs, dvid = dvid)
  bind_rows(doses, obs) |>
    arrange(id, time, desc(evid))
}

# 1. Elderly, 30 ug/kg Q2W x 12 (23 weeks). Dense sampling in the first and
#    seventh dosing intervals (the two the paper compares), plus the trial's
#    own sparse grid over every interval.
ii_q2w   <- 336
elderly_obs <- sort(unique(c(
  seq(0, 336, by = 6),                       # dose 1 interval, dense
  2016 + seq(0, 336, by = 6),                # dose 7 interval, dense
  rep(seq(0, 11) * ii_q2w, each = length(sparse_grid)) + sparse_grid
)))
ev_elderly <- make_events(
  elderly_subj, dose_amt = function(WT) 30 * WT,
  ii = ii_q2w, addl = 11L, obs_times = elderly_obs
) |>
  mutate(cohort = "Elderly 30 ug/kg Q2W")

# 2. Healthy adults, single doses of 10 / 30 / 60 / 120 / 200 ug/kg.
adult_doses <- c(10, 30, 60, 120, 200)
ev_adult <- bind_rows(lapply(seq_along(adult_doses), function(i) {
  d <- adult_doses[i]
  make_events(
    make_adult_subj(n_adult, id_offset = 1000L + (i - 1L) * n_adult),
    dose_amt = function(WT) d * WT, ii = 0, addl = 0L,
    obs_times = sort(unique(c(seq(0, 336, by = 4), sparse_grid)))
  ) |>
    mutate(cohort = sprintf("Adult %d ug/kg SD", d))
}))

events_pk <- bind_rows(ev_elderly, ev_adult)
stopifnot(!anyDuplicated(unique(events_pk[, c("id", "time", "evid", "dvid")])))
```

# Simulation

```{r simulate}
mod <- readModelDb("He_2025_YPEGrhGH")

# useLinCmt = FALSE: rxode2's automatic ODE -> linCmt conversion corrupts the
# dvid -> cmt mapping for multi-endpoint models (see the skill's
# known-vignette-failure-patterns.md pattern 5b).
sim_pk <- rxode2::rxSolve(
  mod, events = events_pk, keep = c("cohort", "AGE", "WT"),
  useLinCmt = FALSE
) |>
  as.data.frame()

stopifnot(nrow(sim_pk) > 0, !anyNA(sim_pk$Cc))
```

For deterministic replication of the published typical-value simulations, the
random effects are zeroed:

```{r simulate-typical}
mod_typical <- rxode2::zeroRe(mod)

# Argument names deliberately differ from the covariate column names so that
# the mutate() below cannot pick up a column instead of the argument.
solve_typical <- function(dose_ug, age_yr, wt_kg, obs_times, ii = 0,
                          addl = 0L) {
  obs <- tibble(
    id = 1L, time = obs_times, amt = NA_real_, evid = 0L, cmt = "central",
    dvid = 1L
  )
  ev <- if (dose_ug > 0) {
    doses <- tibble(
      id = 1L, time = (seq_len(addl + 1L) - 1L) * ii, amt = dose_ug,
      evid = 1L, cmt = "depot", dvid = NA_integer_
    )
    bind_rows(doses, obs)
  } else {
    obs
  }
  ev <- ev |>
    mutate(AGE = age_yr, WT = wt_kg) |>
    arrange(time, desc(evid))
  rxode2::rxSolve(mod_typical, ev, useLinCmt = FALSE) |> as.data.frame()
}
```

## The IGF-1 state starts at its baseline ratio of 1

The PD state is IGF-1 expressed as a ratio to each subject's own baseline, so
the model's typical baseline must be exactly `Kin / Kout = 0.023 / 0.023 = 1`.
This is a deterministic identity, not a cohort statistic, so it is asserted
exactly.

```{r baseline-identity}
base_chk <- solve_typical(0, age_yr = 67.44, wt_kg = 60.54,
                          obs_times = c(0, 24, 336))
stopifnot(
  # No dose -> the IDR sits at its baseline for the whole horizon.
  max(abs(base_chk$IGF1ratio - 1)) < 1e-8
)
base_chk$IGF1ratio
```

# Replicate published figures

## Figure 2C -- elderly versus healthy-adult concentration-time profiles

```{r figure-2c}
sim_pk |>
  filter(time <= 336, cohort %in% c("Elderly 30 ug/kg Q2W", "Adult 30 ug/kg SD",
                                    "Adult 10 ug/kg SD")) |>
  group_by(cohort, time) |>
  summarise(
    Q10 = quantile(Cc, 0.10), Q50 = median(Cc), Q90 = quantile(Cc, 0.90),
    .groups = "drop"
  ) |>
  filter(time > 0) |>
  ggplot(aes(time, Q50, colour = cohort, fill = cohort)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(
    x = "Time after first dose (h)", y = "YPEG-rhGH (ng/mL)",
    colour = NULL, fill = NULL,
    title = "Figure 2C -- elderly 30 ug/kg vs healthy adults",
    caption = paste(
      "Replicates Figure 2C of He 2025: elderly exposure at 30 ug/kg falls",
      "close to the healthy-adult 10 ug/kg profile."
    )
  ) +
  theme(legend.position = "bottom")
```

## Figure 2D -- Cmax and AUC across the healthy-adult dose range

Michaelis-Menten elimination makes exposure rise faster than dose. This is a
typical-value (deterministic) comparison, so it is asserted tightly.

```{r figure-2d}
dose_prop <- bind_rows(lapply(adult_doses, function(d) {
  s <- solve_typical(d * 65.08, age_yr = 30.78, wt_kg = 65.08,
                     obs_times = seq(0, 336, by = 2))
  tibble(
    dose_ug_kg = d,
    cmax = max(s$Cc),
    tmax = s$time[which.max(s$Cc)],
    auclast = sum(diff(s$time) * (head(s$Cc, -1) + tail(s$Cc, -1)) / 2)
  )
}))

knitr::kable(
  dose_prop |>
    mutate(
      dose_norm_cmax = cmax / cmax[dose_ug_kg == 10] / (dose_ug_kg / 10),
      dose_norm_auc  = auclast / auclast[dose_ug_kg == 10] / (dose_ug_kg / 10)
    ) |>
    rename(
      "Dose (ug/kg)"        = dose_ug_kg,
      "Cmax (ng/mL)"        = cmax,
      "Tmax (h)"            = tmax,
      "AUC0-336 (ng*h/mL)"  = auclast,
      "Cmax / dose vs 10 ug/kg" = dose_norm_cmax,
      "AUC / dose vs 10 ug/kg"  = dose_norm_auc
    ),
  digits = 2,
  caption = paste(
    "Typical-value exposure across the healthy-adult dose range. Replicates",
    "the dose-nonlinearity shown in Figure 2D of He 2025."
  )
)

stopifnot(
  # Deterministic: a 20-fold dose increase must give a MORE than 20-fold AUC
  # increase under Michaelis-Menten elimination (Results 3.2.1 / Discussion:
  # "nonlinear elimination characteristics ... in the dose range 10-200 ug/kg").
  dose_prop$auclast[dose_prop$dose_ug_kg == 200] /
    dose_prop$auclast[dose_prop$dose_ug_kg == 10] > 20,
  # Monotone in dose (deterministic, no cohort noise).
  all(diff(dose_prop$cmax) > 0),
  all(diff(dose_prop$auclast) > 0)
)
```

## Figures 9C and 9D -- covariate gradients in the IGF-1 baseline ratio

Figure 9C simulates a 3 mg flat dose Q2W for 52 weeks in 70-year-old elderly
subjects weighing 50, 60, 70 and 80 kg; Figure 9D holds weight at 60 kg and
varies age over 60-80 years. `Emax,72h` is the IGF-1 baseline ratio 72 h after
the last dose and `Etrough` is the ratio immediately before the next dose would
have been due. Both panels are typical-value simulations, so the gradients
below are deterministic.

```{r figure-9cd}
# 26 Q2W doses = 52 weeks; the last dose is at 25 * 336 = 8400 h.
last_dose_time <- 25 * ii_q2w
pd_times <- c(last_dose_time + 72, last_dose_time + ii_q2w)

pd_scenario <- function(age_yr, wt_kg, dose_ug = 3000) {
  s <- solve_typical(dose_ug, age_yr = age_yr, wt_kg = wt_kg,
                     obs_times = pd_times, ii = ii_q2w, addl = 25L)
  tibble(
    AGE = age_yr, WT = wt_kg, dose_mg = dose_ug / 1000,
    Emax_72h = s$IGF1ratio[s$time == pd_times[1]],
    Etrough  = s$IGF1ratio[s$time == pd_times[2]]
  )
}

fig9c <- bind_rows(lapply(c(50, 60, 70, 80), function(w) {
  pd_scenario(age_yr = 70, wt_kg = w)
}))
fig9d <- bind_rows(lapply(c(60, 65, 70, 75, 80), function(a) {
  pd_scenario(age_yr = a, wt_kg = 60)
}))

bind_rows(
  fig9c |> mutate(Panel = "9C: age 70 y, weight varied"),
  fig9d |> mutate(Panel = "9D: weight 60 kg, age varied")
) |>
  select(Panel, AGE, WT, Emax_72h, Etrough) |>
  rename(
    "Panel"          = Panel,
    "Age (years)"    = AGE,
    "Weight (kg)"    = WT,
    "Emax,72h"       = Emax_72h,
    "Etrough"        = Etrough
  ) |>
  knitr::kable(
    digits = 3,
    caption = paste(
      "Replicates Figures 9C and 9D of He 2025 (3 mg Q2W, 52 weeks,",
      "typical-value simulation)."
    )
  )

wt_spread  <- fig9c$Emax_72h[fig9c$WT == 50] - fig9c$Emax_72h[fig9c$WT == 80]
age_spread <- fig9d$Emax_72h[fig9d$AGE == 60] - fig9d$Emax_72h[fig9d$AGE == 80]

stopifnot(
  # Results 3.5: "a significant decrease in the ratio of IGF-1 relative to
  # baseline with increasing body weight". Deterministic typical values, so
  # step-by-step monotonicity is a legitimate assertion here.
  all(diff(fig9c$Emax_72h) < 0),
  all(diff(fig9c$Etrough) < 0),
  # Results 3.5 / Discussion: the age effect is in the same direction but
  # "less" pronounced than the weight effect.
  all(diff(fig9d$Emax_72h) < 0),
  age_spread < wt_spread
)
c(weight_spread = wt_spread, age_spread = age_spread)
```

## Conclusion -- IGF-1 baseline ratio across the 1.5-5 mg flat-dose range

The Conclusion states that "the baseline ratio of administration could rise to
the range of 1.14-1.43 after a 1.5-5 mg dose of 24W/52W given at Q2W in the
elderly, with the 3 mg dose elevating the ratio by approximately 1 SD".
Supplementary Table 6 gives the elderly baseline IGF-1 as mean 93.8 ng/mL with
SD 17.4 ng/mL, so a 1 SD elevation corresponds to a baseline ratio of
`r round(1 + 17.4 / 93.8, 3)`.

```{r conclusion-doses}
sd_ratio <- 1 + 17.4 / 93.8   # Supplementary Table 6: mean 93.8, SD 17.4

flat_dose <- bind_rows(lapply(c(1.5, 3, 5), function(d) {
  pd_scenario(age_yr = 70, wt_kg = 60, dose_ug = d * 1000)
})) |>
  mutate(
    # Mean ratio over the final dosing interval, the quantity the paper's
    # "elevating the ratio by approximately 1 SD" statement is closest to.
    Emean = vapply(dose_mg, function(d) {
      s <- solve_typical(d * 1000, age_yr = 70, wt_kg = 60,
                         obs_times = last_dose_time + seq(0, ii_q2w, by = 6),
                         ii = ii_q2w, addl = 25L)
      mean(s$IGF1ratio)
    }, numeric(1))
  )

flat_dose |>
  select(dose_mg, Emax_72h, Emean, Etrough) |>
  rename(
    "Dose (mg Q2W)"                   = dose_mg,
    "Emax,72h"                        = Emax_72h,
    "Mean over final interval"        = Emean,
    "Etrough"                         = Etrough
  ) |>
  knitr::kable(
    digits = 3,
    caption = paste(
      "IGF-1 baseline ratio over the final dosing interval after 52 weeks of",
      "Q2W dosing (typical 70-year-old, 60 kg). He 2025 Conclusion reports a",
      "range of 1.14-1.43 across 1.5-5 mg, with 3 mg ~ 1 SD =",
      sprintf("%.3f.", sd_ratio)
    )
  )

stopifnot(
  # Deterministic and monotone in dose.
  all(diff(flat_dose$Emax_72h) > 0),
  all(diff(flat_dose$Etrough) > 0),
  # The published 1.14-1.43 range must sit between the model's trough and peak
  # envelopes across 1.5-5 mg (see Assumptions and deviations: the paper does
  # not say which of Emax,72h / Etrough / mean the range refers to).
  min(flat_dose$Etrough) < 1.14,
  max(flat_dose$Emax_72h) > 1.43,
  # The 3 mg interval mean should land near the 1 SD reference. Widened to
  # +/-0.10 around sd_ratio: this is deterministic, but the published claim is
  # qualitative ("approximately 1 SD").
  abs(flat_dose$Emean[flat_dose$dose_mg == 3] - sd_ratio) < 0.10
)
```

## Figure 8 -- IGF-1 baseline-ratio time course in the elderly cohort

```{r figure-8}
pd_obs <- seq(0, 4 * ii_q2w, by = 12)
ev_pd <- bind_rows(
  make_events(elderly_subj, dose_amt = function(WT) 30 * WT, ii = ii_q2w,
              addl = 3L, obs_times = pd_obs, dvid = 2L, cmt_obs = "igf1")
) |>
  mutate(cohort = "Elderly 30 ug/kg Q2W")

sim_pd <- rxode2::rxSolve(mod, events = ev_pd, keep = c("cohort"),
                          useLinCmt = FALSE) |>
  as.data.frame()

sim_pd |>
  group_by(time) |>
  summarise(
    Q10 = quantile(IGF1ratio, 0.10), Q50 = median(IGF1ratio),
    Q90 = quantile(IGF1ratio, 0.90), .groups = "drop"
  ) |>
  ggplot(aes(time / 24, Q50)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.2) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  labs(
    x = "Time (days)", y = "IGF-1 / baseline",
    title = "Figure 8 -- IGF-1 relative to baseline, elderly 30 ug/kg Q2W",
    caption = paste(
      "Replicates the IDR visual predictive check of Figure 8 of He 2025",
      "(median and 10th-90th percentile band of the simulated cohort)."
    )
  )

# The IDR is a stimulation of production, so no subject's ratio ever falls
# below ITS OWN baseline of Kin / Kout_i (the eta on Kout makes that baseline
# subject-specific, so the reference is each subject's t = 0 value rather than
# the typical value of 1). This is a deterministic property of the structure
# and is checked on the residual-error-free IPRED.
pd_baselines <- sim_pd |>
  filter(time == 0) |>
  select(id, baseline = IGF1ratio)
pd_min <- sim_pd |>
  group_by(id) |>
  summarise(min_ratio = min(IGF1ratio), .groups = "drop") |>
  left_join(pd_baselines, by = "id")

# Median profile across the cohort, used for the TEmax-lags-Tmax check below.
pd_median <- tapply(sim_pd$IGF1ratio, sim_pd$time, median)
pd_temax <- as.numeric(names(pd_median)[which.max(pd_median)])

stopifnot(
  nrow(pd_min) == n_elderly,
  all(pd_min$min_ratio >= pd_min$baseline - 1e-8),
  # Results 3.3 / Discussion: "TEmax ... was significantly longer than the Tmax
  # of YPEG-rhGH", with a median TEmax of 72-96 h. Asserted only as "later than
  # one day", which holds for any cohort the model can draw.
  pd_temax > 24
)
pd_temax
```

# PKNCA validation

## Healthy-adult single-dose NCA by dose group

```{r pknca-adult}
adult_nca_conc <- sim_pk |>
  filter(grepl("^Adult", cohort), !is.na(Cc)) |>
  select(id, time, Cc, cohort)

# Guarantee a time = 0 record per subject (extravascular pre-dose Cc = 0).
adult_nca_conc <- bind_rows(
  adult_nca_conc,
  adult_nca_conc |> distinct(id, cohort) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, cohort, time, .keep_all = TRUE) |>
  arrange(id, time)

adult_conc_obj <- PKNCA::PKNCAconc(adult_nca_conc, Cc ~ time | cohort + id)

adult_dose_df <- ev_adult |>
  filter(evid == 1) |>
  select(id, time, amt, cohort)
adult_dose_obj <- PKNCA::PKNCAdose(adult_dose_df, amt ~ time | cohort + id)

adult_intervals <- data.frame(
  start = 0, end = 336,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, half.life = TRUE
)

adult_nca <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(adult_conc_obj, adult_dose_obj, intervals = adult_intervals)
)

adult_summary <- as.data.frame(adult_nca) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "half.life")) |>
  group_by(cohort, PPTESTCD) |>
  summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = PPTESTCD, values_from = median)

adult_summary |>
  rename(
    "Dose group"          = cohort,
    "Cmax (ng/mL)"        = cmax,
    "Tmax (h)"            = tmax,
    "AUC0-336 (ng*h/mL)"  = auclast,
    "t1/2 (h)"            = half.life
  ) |>
  knitr::kable(
    digits = 2,
    caption = "PKNCA summary of the simulated healthy-adult single-dose cohorts."
  )
```

## Elderly accumulation between the first and seventh dosing intervals

The paper reports accumulation ratios of 1.34 (Cmax), 1.35 (Cmin) and 1.23
(AUC) for the elderly cohort, and states that steady state was reached by
week 13. The seventh injection (week 12, `t = 2016 h`) is the interval the
paper's week-13 sampling covers.

```{r pknca-elderly}
elderly_nca_conc <- sim_pk |>
  filter(cohort == "Elderly 30 ug/kg Q2W", !is.na(Cc)) |>
  select(id, time, Cc, cohort)
elderly_nca_conc <- bind_rows(
  elderly_nca_conc,
  elderly_nca_conc |> distinct(id, cohort) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, cohort, time, .keep_all = TRUE) |>
  arrange(id, time)

elderly_conc_obj <- PKNCA::PKNCAconc(elderly_nca_conc, Cc ~ time | cohort + id)
elderly_dose_obj <- PKNCA::PKNCAdose(
  ev_elderly |> filter(evid == 1) |> select(id, time, amt, cohort),
  amt ~ time | cohort + id
)

# clast.obs, not cmin: PKNCA's cmin over an interval is the minimum anywhere in
# it, which for a Q2W interval is the trough at its START. The trough the paper
# reports is the concentration at the END of the interval.
elderly_intervals <- data.frame(
  start = c(0, 2016), end = c(336, 2352),
  cmax = TRUE, tmax = TRUE, auclast = TRUE, clast.obs = TRUE
)

elderly_nca <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(elderly_conc_obj, elderly_dose_obj,
                   intervals = elderly_intervals)
)

elderly_res <- as.data.frame(elderly_nca) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "clast.obs")) |>
  mutate(interval = ifelse(start == 0, "dose1", "dose7")) |>
  select(id, PPTESTCD, interval, PPORRES) |>
  pivot_wider(names_from = interval, values_from = PPORRES) |>
  mutate(ratio = dose7 / dose1)

accum <- elderly_res |>
  filter(PPTESTCD %in% c("cmax", "clast.obs", "auclast")) |>
  group_by(PPTESTCD) |>
  summarise(median_ratio = median(ratio, na.rm = TRUE), .groups = "drop") |>
  mutate(
    published = c(auclast = 1.23, clast.obs = 1.35, cmax = 1.34)[PPTESTCD],
    pct_diff = 100 * (median_ratio - published) / published
  )

accum |>
  rename(
    "NCA parameter"           = PPTESTCD,
    "Simulated dose7 / dose1" = median_ratio,
    "He 2025 reported"        = published,
    "Difference (%)"          = pct_diff
  ) |>
  knitr::kable(
    digits = 3,
    caption = paste(
      "Accumulation from the first to the seventh Q2W dosing interval,",
      "elderly cohort. He 2025 Results 3.2 reports RCmax 1.34, RCmin 1.35",
      "and RAUC 1.23."
    )
  )

stopifnot(
  # Results 3.2: "a moderate increase in serum exposure ... but no significant
  # drug accumulation". The bound admits any cohort the model can draw while
  # still breaking on a mis-transcribed dosing interval or clearance term
  # (which would move accumulation well outside 1.0-1.7).
  all(accum$median_ratio > 1.0),
  all(accum$median_ratio < 1.7)
)
```

## Comparison against published NCA

The paper reports no absolute NCA table; the only NCA values it publishes are
the elderly median Tmax after the first dose (46.16 h) and after later doses
(22.71 h). Those were measured on the trial's own sparse sampling grid
(0, 8, 24, 72, 96, 120, 168 and 336 h after a dose, Supplementary Table 3), on
which an observed Tmax can only land on one of eight times. The simulated Tmax
below is therefore computed on that same grid.

```{r nca_compare}
sparse_nca_conc <- sim_pk |>
  filter(
    cohort == "Elderly 30 ug/kg Q2W",
    time %in% c(sparse_grid, 2016 + sparse_grid)
  ) |>
  mutate(
    treatment = ifelse(time <= 336, "Elderly, dose 1", "Elderly, dose 7")
  ) |>
  select(id, time, Cc, treatment)

sparse_conc_obj <- PKNCA::PKNCAconc(sparse_nca_conc, Cc ~ time | treatment + id)
sparse_dose_obj <- PKNCA::PKNCAdose(
  ev_elderly |>
    filter(evid == 1, time %in% c(0, 2016)) |>
    mutate(treatment = ifelse(time == 0, "Elderly, dose 1",
                              "Elderly, dose 7")) |>
    select(id, time, amt, treatment),
  amt ~ time | treatment + id
)

sparse_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  sparse_conc_obj, sparse_dose_obj,
  intervals = data.frame(
    start = c(0, 2016), end = c(336, 2352), cmax = TRUE, tmax = TRUE
  )
))

published <- tibble::tribble(
  ~treatment,        ~tmax,
  "Elderly, dose 1", 46.16,
  "Elderly, dose 7", 22.71
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sparse_nca,
  reference = published,
  by        = "treatment",
  params    = "tmax",
  units     = c(tmax = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated vs published median Tmax on the trial's own sampling grid.",
    "* differs from the reference by more than 20%."
  )
)
```

The dose-1 row is starred: the model's typical absorption
(`Ka = 0.01086 * (AGE/35.5)^-0.6957` = `r sprintf("%.5f", 0.01086 * (67.44 / 35.5)^-0.6957)` /h at
the cohort's mean age) peaks earlier than the observed median. This is recorded
as a deviation below rather than tuned away.

The underlying finding the paper draws from that Tmax -- "the Tmax remained
slower compared to younger subjects", "markedly reduced absorption in the
elderly population" -- *is* reproduced, and is checked directly on the
individual `Ka` values rather than on a Tmax read off two different sampling
grids:

```{r ka-age-direction}
ka_by_cohort <- sim_pk |>
  filter(cohort %in% c("Elderly 30 ug/kg Q2W", "Adult 30 ug/kg SD")) |>
  group_by(cohort, id) |>
  summarise(ka = first(ka), .groups = "drop") |>
  group_by(cohort) |>
  summarise(
    median_ka = median(ka),
    median_abs_half_life = log(2) / median(ka),
    .groups = "drop"
  )

ka_by_cohort |>
  rename(
    "Cohort"                       = cohort,
    "Median Ka (1/h)"              = median_ka,
    "Absorption half-life (h)"     = median_abs_half_life
  ) |>
  knitr::kable(
    digits = c(0, 5, 1),
    caption = paste(
      "Individual absorption rate constants by cohort. The AGE exponent on Ka",
      "is negative (-0.6957), so the elderly absorb more slowly."
    )
  )

ka_elderly <- ka_by_cohort$median_ka[
  ka_by_cohort$cohort == "Elderly 30 ug/kg Q2W"
]
ka_adult <- ka_by_cohort$median_ka[
  ka_by_cohort$cohort == "Adult 30 ug/kg SD"
]
stopifnot(
  # Ka has no eta beyond the fixed omega and the age ranges of the two cohorts
  # do not overlap, so this is a large, robust effect (roughly 1.7-fold) rather
  # than a race between two noisy statistics.
  ka_elderly < 0.8 * ka_adult
)
```

## Elderly versus healthy-adult exposure at the same 30 ug/kg dose

Results 3.2 reports that elderly Cmax and AUC0-last at 30 ug/kg reached only
7.75% and 19.13% of the healthy-adult values at the same dose. The model
reproduces the direction and a substantial magnitude, but not the full size of
the observed gap.

```{r elderly-vs-adult}
exposure <- as.data.frame(adult_nca) |>
  filter(cohort == "Adult 30 ug/kg SD", PPTESTCD %in% c("cmax", "auclast")) |>
  group_by(PPTESTCD) |>
  summarise(adult = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  left_join(
    elderly_res |>
      filter(PPTESTCD %in% c("cmax", "auclast")) |>
      group_by(PPTESTCD) |>
      summarise(elderly = median(dose1, na.rm = TRUE), .groups = "drop"),
    by = "PPTESTCD"
  ) |>
  mutate(
    pct_of_adult = 100 * elderly / adult,
    published_pct = c(auclast = 19.13, cmax = 7.75)[PPTESTCD]
  )

exposure |>
  rename(
    "NCA parameter"          = PPTESTCD,
    "Adult 30 ug/kg (median)" = adult,
    "Elderly 30 ug/kg (median)" = elderly,
    "Elderly as % of adult"  = pct_of_adult,
    "He 2025 observed %"     = published_pct
  ) |>
  knitr::kable(
    digits = 2,
    caption = paste(
      "Elderly exposure as a percentage of healthy-adult exposure at the same",
      "30 ug/kg dose. He 2025 Results 3.2 reports 7.75% (Cmax) and 19.13%",
      "(AUC0-last) from observed data."
    )
  )

stopifnot(
  # DIRECTIONAL gate only -- the magnitude is a documented deviation (below).
  # Elderly exposure must be well under two thirds of the adult value at the
  # same dose; the AGE effects on Ka, V1/F and Vmax make this a large, robust
  # effect. Realised 23.4% (Cmax) and 46.4% (AUC) at 16 threads. Dropping the
  # AGE terms would push both toward 100%, so 65 still goes red.
  all(exposure$pct_of_adult < 65)
)
```

# Assumptions and deviations

* **Covariate centring values come from the supplement, not the main text.**
  Equation 1 defines the covariate model as
  `theta_i = theta_TV x (Cov_i / Cov_median)^theta_x` but never prints
  `Cov_median`. The supplement's "PopPK model equations" section gives the
  final equations with the centring values substituted (`AGE / 35.5` and
  `WEIGHT / 70`), and those are what the model file uses. 35.5 years is not
  tabulated anywhere as a summary statistic; it is consistent with the median
  age of the pooled 52-subject dataset (36 adults aged 21-44 plus 16 elderly
  aged 65-74, Table 2). 70 kg is a rounded reference, not the dataset median
  (59.0 kg elderly / 66.4 kg adults).
* **`omega(Ka)` and `omega(Km)` are encoded as fixed.** Supplementary Table 1
  reports both as exactly 54.77% with no RSE and no confidence interval in the
  final-model column *and* in the 1,000-replicate bootstrap column. A parameter
  estimated freely cannot reproduce to four significant figures across 1,000
  bootstrap refits, so both were held constant. `0.5477^2 = 0.300`, a round
  variance consistent with a held initial estimate; they are encoded as
  `fixed(0.3)`.
* **The IIV percentages are read as `omega x 100`, not as
  `sqrt(exp(omega^2) - 1) x 100`.** The `0.5477 -> 0.300` coincidence above
  only holds under the first convention. The paper describes the retained PD
  random effects as "additive random effects for EC50 and Kout"; taken
  literally with an additive eta, `omega(EC50) = 78.76` would be an SD of 0.79
  on an EC50 of 50.74 ng/mL, i.e. essentially no variability, which is
  irreconcilable with the reported 9.97% shrinkage and 30.66% RSE. They are
  therefore encoded as log-normal (exponential) random effects, which is what
  the reported percentages and shrinkages are consistent with.
* **Residual-error scales.** Both tables print the residual point estimate on a
  percentage-style scale while the main text's 95% CI column prints the same
  quantity as a fraction (PK: 28.94 vs 0.2730-0.3057; PD: 13.20 vs
  0.1161-0.1479; the supplement's bootstrap tables give 27.30-30.57 and
  11.61-14.79). Both are entered as fractions: `propSd = 0.2894` and
  `addSd_IGF1ratio = 0.1320`. An additive residual SD of 13.20 on a
  ratio-to-baseline scale that runs from 1 to about 2 would be nonsensical.
* **`Kin` carries the printed unit `ng/(mL*h)` but is dimensionless-per-hour.**
  The PD variable the model fits is the IGF-1 *ratio to baseline*, so `Kin` and
  `Kout` are both `1/h` on that scale and their ratio `Kin/Kout = 1` is the
  baseline. The `ng/(mL*h)` tag in Table 4 is carried over from the absolute
  IGF-1 scale.
* **The PD parameters are the elderly-subject estimates.** Table 4 is headed
  "PK/PD final model parameters of IGF-1 relative to baseline ratio in elderly
  subjects", and the PD layer was fitted sequentially with individual PK
  parameters taken from the final PK model. Supplementary Figures 4 and 5 show
  that the same IDR structure was also fitted in healthy adults, but those
  estimates are not published, so the file carries only the elderly values.
* **No bioavailability term.** `V1/F`, `V2/F` and `CL2/F` are apparent
  parameters throughout; `F` was not estimated, so no `f(depot)` is applied and
  the doses in the vignette are the administered subcutaneous amounts.
* **Cohort covariate distributions are assumed normal**, truncated at the
  minima and maxima of source Table 2. The paper reports only mean, SD, median
  and range, not the full distributions. Sex is not simulated because it is not
  a model covariate.
* **DEVIATION -- the elderly/adult exposure gap is under-reproduced.** He 2025
  reports observed elderly Cmax and AUC0-last at 30 ug/kg as 7.75% and 19.13%
  of the healthy-adult values; the model gives roughly 23% and 46%. The
  published percentages are non-compartmental summaries of observed
  concentrations from an eight-point sampling grid in a 16-subject cohort with
  97% CV on `V1/F`, and the model is the shrunken population fit to the pooled
  data. The direction and a large fraction of the magnitude are reproduced; the
  gate above therefore checks only that elderly exposure is well under half the
  adult value. The published values are not reproducible from the published
  parameters and are recorded here rather than tuned to.
* **DEVIATION -- median Tmax after the first elderly dose.** The paper reports
  46.16 h; the model gives 24 h on the same eight-point sampling grid
  (about 18 h on a dense grid, typical value). With `Ka = 0.0070 /h` at the
  cohort mean age the absorption half-life is about 100 h, so the profile is
  flip-flop and the peak is broad and flat -- observed Tmax on a grid whose
  neighbouring points are 24 h and 72 h is very sensitive to assay noise, which
  the simulated `Cc` (an IPRED) excludes. The starred row in the NCA comparison
  table records the difference. The paper's *later*-dose Tmax of 22.71 h is
  reproduced closely (24 h, +5.7%), so the disagreement is confined to the
  first-dose value.
* **The Conclusion's 1.14-1.43 range is bracketed, not matched exactly.** The
  paper does not state whether that range refers to peak, trough or mean
  baseline ratio, nor at which age and weight. At a typical 70-year-old, 60 kg
  subject the model gives troughs of 1.06-1.19 and 72 h peaks of 1.21-1.66 over
  1.5-5 mg Q2W; the published range sits inside that envelope, and the 3 mg
  interval mean is close to the 1 SD reference of
  `r round(1 + 17.4 / 93.8, 3)` derived from Supplementary Table 6.
* **No non-paper-derived parameter values.** Every `ini()` entry is traceable to
  the article's Table 3, Table 4, Results text, or the open-access
  Supplementary Material; nothing was digitised from a figure or supplied by
  correspondence.
