---
title: "Wuzhi-capsule lignans as CYP3A4/CYP3A5 inhibitors (He 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Wuzhi-capsule lignans as CYP3A4/CYP3A5 inhibitors (He 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("He_2021_schisantherinA_cyp3a"))$meta$reference`
* Article: <https://doi.org/10.3390/ph14030198>
* PubMed Central open-access copy: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7997453/>

He 2021 co-administer the Wuzhi capsule (a standardised ethanolic extract of
*Schisandra sphenanthera*) with tacrolimus, a combination used in Chinese
transplant practice because the extract's lignans inhibit CYP3A and raise
tacrolimus exposure, cutting the required tacrolimus dose. This paper measures,
for the two most abundant lignans, schisantherin A (STA) and schisandrin A
(SIA), their reversible and time-dependent inhibition of CYP3A4 and CYP3A5
separately, using CYP3A5-genotyped human liver microsomes with CYP3cide to
isolate CYP3A5.

This paper contributes two model files, one per lignan:

```{r models}
mod_sta <- rxode2::rxode2(readModelDb("He_2021_schisantherinA_cyp3a"))
mod_sia <- rxode2::rxode2(readModelDb("He_2021_schisandrinA_cyp3a"))

# Fixed and estimated ini() values are not returned as columns by rxSolve(), so
# read them straight off the packaged model when a check needs one.
ini_val <- function(mod, nm) {
  stopifnot(nm %in% mod$iniDf$name)
  mod$iniDf$est[mod$iniDf$name == nm]
}
```

# Scope: what this paper does and does not contribute

He 2021 has two halves, and only one of them is reproducible.

The **first half is the authors' own bench work**: reversible-inhibition (RI)
and time-dependent-inhibition (TDI) assays that measure the inhibitory potency
of STA and SIA on CYP3A4 and CYP3A5, each isoform in its own microsomal system.
Every fitted constant is printed, the assay designs are in the Methods, and the
underlying data points are in Supplementary Table S1. That is the layer this
vignette validates and that the two model files carry.

The **second half is a Simcyp V13.1.1 whole-body physiologically based model**
that integrates those constants (together with the enzyme abundances of Table 4
and the tacrolimus clearance kinetics of Table 3) to predict the fold-increase
in tacrolimus AUC in CYP3A5 expressers and non-expressers. That layer is a
vendor platform model and is **not** extracted: the whole-body ODE system is
Simcyp's rather than the authors', the tissue Kp values were screened and
optimised inside the platform, and no project file is deposited. Reconstructing
it would require substituting physiology from outside the source. The section
"Published results not reproduced here" records what that half produced.

A load-bearing consequence for the in vitro layer: the two isoforms are
inhibited by different combinations of the two mechanisms, and each model file
records exactly the combination the paper fitted.

```{r mechanism_table, echo = FALSE}
tibble::tibble(
  Lignan = c("Schisantherin A (STA)", "Schisantherin A (STA)",
             "Schisandrin A (SIA)", "Schisandrin A (SIA)"),
  Isoform = c("CYP3A4", "CYP3A5", "CYP3A4", "CYP3A5"),
  `Reversible (Ki, uM)` = c("0.15", "0.11", "little (not fitted)", "8.74"),
  `Time-dependent (kinact /min, KI uM)` = c("0.11, 2.45", "none (Figure 3C)",
                                            "0.019, 2.54", "0.014, 2.07")
) |>
  knitr::kable(caption = "Inhibition mechanisms fitted by He 2021 (Results 2.1, 2.2; Figures 2-4).")
```

# In vitro systems

```{r systems, echo = FALSE}
tibble::tibble(
  Assay = c("Reversible inhibition (RI)", "Time-dependent inhibition (TDI)"),
  `CYP3A4 system` = c("CYP3A5*3/*3 HLM, 0.2 mg/mL", "CYP3A5*3/*3 HLM, 0.5 mg/mL"),
  `CYP3A5 system` = c("CYP3A5*1/*3 HLM + 0.5 uM CYP3cide, 0.2 mg/mL",
                      "CYP3A5*1/*3 HLM + 1.2 uM CYP3cide, 0.5 mg/mL"),
  Probe = c("Tacrolimus disappearance (Dixon plot)",
            "Testosterone 6-beta-hydroxylation (double-reciprocal plot)"),
  Buffer = c("0.1 M potassium phosphate pH 7.4, 1 mM NADPH, 37 C",
             "0.1 M PBS pH 7.4, 1 mM NADPH, 37 C")
) |>
  knitr::kable(caption = "In vitro systems of He 2021 (Methods 4.3, 4.4).")
```

CYP3cide (PF-04981517) is a selective mechanism-based inactivator of CYP3A4; it
is used to knock out CYP3A4 so that the residual testosterone or tacrolimus
turnover in the CYP3A5*1/*3 microsomes is attributable to CYP3A5.

# Source trace

Every value in the two model files, with the location it came from.

```{r source_trace, echo = FALSE}
tibble::tribble(
  ~Model,                              ~Quantity,        ~Value,        ~`Source location`,
  "He_2021_schisantherinA_cyp3a",      "ki_3a4",         "0.15 uM",     "Figure 2A annotation; Results 2.1",
  "He_2021_schisantherinA_cyp3a",      "ki_3a5",         "0.11 uM",     "Figure 2B annotation; Results 2.1",
  "He_2021_schisantherinA_cyp3a",      "ki_inact_3a4",   "2.45 uM",     "Figure 3B annotation; Discussion",
  "He_2021_schisantherinA_cyp3a",      "kinact_3a4",     "0.11 /min",   "Figure 3B annotation; Discussion",
  "He_2021_schisantherinA_cyp3a",      "(CYP3A5 TDI)",   "none",        "Figure 3C (no inactivation observed)",
  "He_2021_schisandrinA_cyp3a",        "ki_3a5",         "8.74 uM",     "Figure 2D annotation; Results 2.1",
  "He_2021_schisandrinA_cyp3a",        "(CYP3A4 RI)",    "little",      "Figure 2C (no significant reversible inhibition)",
  "He_2021_schisandrinA_cyp3a",        "ki_inact_3a4",   "2.54 uM",     "Figure 4B annotation; Discussion",
  "He_2021_schisandrinA_cyp3a",        "kinact_3a4",     "0.019 /min",  "Figure 4B annotation; Discussion",
  "He_2021_schisandrinA_cyp3a",        "ki_inact_3a5",   "2.07 uM",     "Figure 4D annotation; Discussion",
  "He_2021_schisandrinA_cyp3a",        "kinact_3a5",     "0.014 /min",  "Figure 4D annotation; Discussion",
  "both",                              "addSd (fixed 0)","0",           "Not reported anywhere; see Assumptions",
  "Inactivation-rate form",            "kobs = kinact*I/(KI+I)", "-",   "Methods 4.4, Equation (1)",
  "Inactivation ODE",                  "d/dt(E) = -kobs*E",     "-",    "log-linear decay of Figures 3A, 4A, 4C",
  "Reversible form",                   "1/(1 + I/Ki)",          "-",    "Competitive inhibition, Dixon plots (Figure 2)"
) |>
  knitr::kable(caption = "Source trace for every ini() value and every model equation.")
```

The eight fitted constants live in the annotations of Figures 2-4 (and are
repeated in the Results and Discussion text); they were read from the rendered
figure panels, which is why a PDF-to-text pass does not surface all of them.
They are the authors' own reported estimates, not digitised from the plotted
points.

# Structural identities of the inactivation model

The time-dependent inactivation is a first-order decay whose rate constant
`kobs` is a hyperbolic (Michaelis-Menten) function of the inhibitor
concentration. These identities are exact by construction and pin the
transcription of `kinact` and `KI`.

```{r identities}
kobs <- function(mod, isoform, I) {
  kinact <- ini_val(mod, paste0("kinact_", isoform))
  KI     <- ini_val(mod, paste0("ki_inact_", isoform))
  kinact * I / (KI + I)
}

# 1. kobs(0) = 0 exactly (no inhibitor => no inactivation).
stopifnot(kobs(mod_sta, "3a4", 0) == 0)
stopifnot(kobs(mod_sia, "3a4", 0) == 0)
stopifnot(kobs(mod_sia, "3a5", 0) == 0)

# 2. kobs(KI) = kinact/2 exactly (half-maximal at I = KI).
for (arg in list(list(mod_sta, "3a4"), list(mod_sia, "3a4"), list(mod_sia, "3a5"))) {
  m <- arg[[1]]; iso <- arg[[2]]
  KI <- ini_val(m, paste0("ki_inact_", iso))
  kinact <- ini_val(m, paste0("kinact_", iso))
  stopifnot(abs(kobs(m, iso, KI) - kinact / 2) < 1e-12)
}

# 3. Plateau: kobs approaches kinact at saturating I.
stopifnot(abs(kobs(mod_sta, "3a4", 1e6) - ini_val(mod_sta, "kinact_3a4")) < 1e-6)
```

# Replicating the inactivation time courses (Figures 3A, 4A, 4C)

The ODE `d/dt(E) = -kobs * E` with `E(0) = 1` has the closed-form solution
`E(t) = exp(-kobs * t)`, which is the straight line the authors plot on a log
axis. The model is solved over the assay's preincubation grid at each measured
inhibitor concentration.

```{r decay_sim}
solve_decay <- function(mod, isoform, conc, covname, tmax = 30) {
  ev <- do.call(rbind, lapply(seq_along(conc), function(i) {
    d <- data.frame(id = i, time = seq(0, tmax, by = 1), evid = 0L, cmt = paste0("enzyme_", isoform))
    d[[covname]] <- conc[i]
    d
  }))
  state <- paste0("enzyme_", isoform)
  out <- rxode2::rxSolve(mod, ev, returnType = "data.frame")
  out$conc <- conc[out$id]
  out$activity <- out[[state]]
  out
}

# STA on CYP3A4 (Figure 3A): I = 0, 0.25, 0.5, 1, 2 uM.
sta_3a4 <- solve_decay(mod_sta, "3a4", c(0, 0.25, 0.5, 1, 2), "CP_STA_UM")

# SIA on CYP3A4 (Figure 4A) and CYP3A5 (Figure 4C): I = 0, 2, 4, 8, 16 uM.
sia_3a4 <- solve_decay(mod_sia, "3a4", c(0, 2, 4, 8, 16), "CP_SIA_UM")
sia_3a5 <- solve_decay(mod_sia, "3a5", c(0, 2, 4, 8, 16), "CP_SIA_UM")
```

```{r decay_check}
# The ODE solution must equal exp(-kobs*t) to solver tolerance.
check_closed_form <- function(df, mod, isoform) {
  df$kobs <- kobs(mod, isoform, df$conc)
  df$analytic <- exp(-df$kobs * df$time)
  max(abs(df$activity - df$analytic))
}
stopifnot(check_closed_form(sta_3a4, mod_sta, "3a4") < 1e-5)
stopifnot(check_closed_form(sia_3a4, mod_sia, "3a4") < 1e-5)
stopifnot(check_closed_form(sia_3a5, mod_sia, "3a5") < 1e-5)
```

```{r decay_plot, echo = FALSE, fig.cap = "Modelled log-percent CYP activity remaining during preincubation. Replicates the linear-regression lines of He 2021 Figures 3A (STA on CYP3A4), 4A (SIA on CYP3A4) and 4C (SIA on CYP3A5)."}
plot_df <- dplyr::bind_rows(
  dplyr::mutate(sta_3a4, panel = "STA on CYP3A4 (Fig 3A)"),
  dplyr::mutate(sia_3a4, panel = "SIA on CYP3A4 (Fig 4A)"),
  dplyr::mutate(sia_3a5, panel = "SIA on CYP3A5 (Fig 4C)")
)
ggplot(plot_df, aes(time, log(100 * activity), colour = factor(conc), group = conc)) +
  geom_line() +
  facet_wrap(~panel) +
  labs(x = "Preincubation time (min)", y = "ln(% activity of control)", colour = "I (uM)") +
  theme_bw()
```

# Cross-check against the Supplementary Table S1 data points

Table S1 gives the raw natural-log percent-activity readings behind Figures 3A,
4A and 4C. Regressing each column against preincubation time recovers an
observed `kobs` per concentration; these must agree with the model's
`kinact*I/(KI+I)` prediction. Because these are experimental means with the
authors' own scatter, the check is an envelope on the centre and spread of the
relative difference, not an exact bound.

```{r s1_check}
tt <- c(30, 20, 10, 5, 0)
# He 2021 Table S1, Figure 3A (STA on CYP3A4), ln(% activity) at I = 2, 1, 0.5, 0.25 uM.
s1_sta_3a4 <- list(
  I = c(2, 1, 0.5, 0.25),
  m = cbind(
    c(3.0051, 3.1694, 3.5615, 4.0322, 4.6052),
    c(3.5554, 3.8785, 4.0515, 4.1658, 4.5902),
    c(3.9183, 3.9359, 4.0778, 4.2393, 4.6010),
    c(4.2407, 4.2986, 4.3386, 4.4164, 4.6010)
  )
)
# Table S1, Figure 4A (SIA on CYP3A4), I = 16, 8, 4, 2 uM.
s1_sia_3a4 <- list(
  I = c(16, 8, 4, 2),
  m = cbind(
    c(4.0738, 4.2397, 4.3776, 4.4518, 4.6052),
    c(4.1876, 4.2781, 4.3979, 4.5164, 4.6029),
    c(4.2490, 4.3925, 4.4420, 4.5505, 4.6052),
    c(4.3520, 4.4602, 4.5165, 4.5754, 4.6052)
  )
)
# Table S1, Figure 4C (SIA on CYP3A5), I = 16, 8, 4, 2 uM.
s1_sia_3a5 <- list(
  I = c(16, 8, 4, 2),
  m = cbind(
    c(4.1607, 4.2585, 4.3701, 4.4523, 4.5539),
    c(4.2282, 4.4028, 4.4150, 4.4632, 4.6052),
    c(4.2859, 4.4869, 4.5031, 4.4998, 4.6052),
    c(4.3791, 4.4427, 4.5218, 4.5539, 4.6052)
  )
)

observed_kobs <- function(s1) {
  vapply(seq_along(s1$I), function(j) unname(-coef(lm(s1$m[, j] ~ tt))[2]), numeric(1))
}

compare_kobs <- function(s1, mod, isoform) {
  obs <- observed_kobs(s1)
  pred <- kobs(mod, isoform, s1$I)
  100 * (obs - pred) / pred
}

pct_sta_3a4 <- compare_kobs(s1_sta_3a4, mod_sta, "3a4")
pct_sia_3a4 <- compare_kobs(s1_sia_3a4, mod_sia, "3a4")
pct_sia_3a5 <- compare_kobs(s1_sia_3a5, mod_sia, "3a5")

pct_all <- c(pct_sta_3a4, pct_sia_3a4, pct_sia_3a5)
summary(pct_all)

stopifnot(
  # Centre: a mis-transcribed kinact or KI shifts every column together.
  abs(median(pct_all)) < 10,
  # Envelope: robust to the assay scatter in individual columns.
  quantile(abs(pct_all), 0.9) < 25
)
```

```{r s1_table, echo = FALSE}
tibble::tibble(
  Panel = c(rep("STA CYP3A4 (Fig 3A)", 4), rep("SIA CYP3A4 (Fig 4A)", 4), rep("SIA CYP3A5 (Fig 4C)", 4)),
  `I (uM)` = c(s1_sta_3a4$I, s1_sia_3a4$I, s1_sia_3a5$I),
  `Observed kobs (/min)` = round(c(observed_kobs(s1_sta_3a4), observed_kobs(s1_sia_3a4), observed_kobs(s1_sia_3a5)), 4),
  `Model kobs (/min)` = round(c(kobs(mod_sta, "3a4", s1_sta_3a4$I), kobs(mod_sia, "3a4", s1_sia_3a4$I), kobs(mod_sia, "3a5", s1_sia_3a5$I)), 4),
  `Diff (%)` = round(pct_all, 1)
) |>
  knitr::kable(caption = "Observed inactivation rate constants (regressed from Table S1) vs the fitted hyperbola.")
```

# Reversible-inhibition factors

The Dixon-plot inhibition constants enter a victim-drug clearance as the
competitive factor `1/(1 + I/Ki)`, exposed by each model as `riFactor_*`. At
`I = Ki` the factor is exactly one half, and STA's much smaller CYP3A5 Ki
(0.11 uM) than SIA's (8.74 uM) makes it the far stronger reversible inhibitor.

```{r reversible}
ri_factor <- function(I, Ki) 1 / (1 + I / Ki)

# Exact identity at I = Ki.
stopifnot(abs(ri_factor(ini_val(mod_sta, "ki_3a5"), ini_val(mod_sta, "ki_3a5")) - 0.5) < 1e-12)
stopifnot(abs(ri_factor(ini_val(mod_sia, "ki_3a5"), ini_val(mod_sia, "ki_3a5")) - 0.5) < 1e-12)

# STA is the stronger reversible CYP3A5 inhibitor at a common concentration.
I_common <- 0.5
sta_loss <- 1 - ri_factor(I_common, ini_val(mod_sta, "ki_3a5"))
sia_loss <- 1 - ri_factor(I_common, ini_val(mod_sia, "ki_3a5"))
stopifnot(sta_loss > sia_loss)
c(STA_CYP3A5_activity_loss = sta_loss, SIA_CYP3A5_activity_loss = sia_loss)
```

# Published results not reproduced here

The Simcyp whole-body half predicted the fold-increase (AUC ratio, AUCR) in
tacrolimus exposure when co-dosed with each lignan, split by CYP3A5 genotype and
by which inhibition mechanism was switched on (Table 2). For a multidose of STA
the predicted AUCR was 2.70 in CYP3A5 expressers and 2.41 in non-expressers; SIA
raised exposure less. These numbers depend on the enzyme abundances of Table 4,
the tacrolimus CYP kinetics of Table 3, the tissue-partition screen and the
Simcyp physiology library, none of which is an nlmixr2 ODE model, so they are
recorded here rather than reproduced.

# Assumptions and deviations

* **No residual-error model.** The source fits the RI and TDI assays by linear
  and nonlinear regression and reports only the resulting constants; there is no
  residual-error model and no assay CV. Per the library convention the additive
  residual SD is fixed at zero, so each model returns the deterministic
  published curve.
* **No inter-individual variability.** The reported `+/-` terms are standard
  errors of the regression, not between-donor variability, so they are recorded
  in `population` and not encoded as an omega. Pooled genotyped microsomes are a
  single system, not a population.
* **Constants read from figure annotations.** The eight fitted constants are
  printed in the annotations of Figures 2-4 (and repeated in the Results and
  Discussion text); they were read from the rendered panels rather than a table.
  They are the authors' own estimates.
* **kinact time unit.** `kinact` is carried in per-minute units, matching the
  per-minute preincubation-time axis of the assays; the model time unit is
  minutes.
* **No enzyme turnover.** The CYP degradation rate constant `kdeg`, which
  governs the in-vivo magnitude of TDI, is a Simcyp default and is not printed.
  No turnover term is carried, so the models describe the inactivation phase of
  the in-vitro assay only; resynthesis is negligible over the 30-minute
  incubation.
* **Mechanisms carried per isoform as fitted.** STA has no time-dependent
  inactivation of CYP3A5 (Figure 3C), so that state has no inactivation term;
  SIA has little reversible inhibition of CYP3A4 (Figure 2C), so that model
  carries no CYP3A4 reversible constant. Each file records only the mechanism
  the paper actually fitted for each isoform.
* **In-vivo use requires an unbound driving concentration.** Applying either
  model to a plasma profile requires supplying the unbound inhibitor
  concentration at the site of enzyme interaction as `CP_STA_UM` / `CP_SIA_UM`;
  the in-vitro incubation concentration is not a plasma concentration.
