---
title: "Biologics and small molecules for psoriasis: PASI75 and PASI90 MBMA (He 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Biologics and small molecules for psoriasis: PASI75 and PASI90 MBMA (He 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

He H, Wu W, Zhang Y, Zhang M, Sun N, Zhao L, Wang X. Model-Based Meta-Analysis
in Psoriasis: A Quantitative Comparison of Biologics and Small Targeted
Molecules. *Front Pharmacol.* 2021;12:586827.
[doi:10.3389/fphar.2021.586827](https://doi.org/10.3389/fphar.2021.586827).

He 2021 is a longitudinal model-based meta-analysis (MBMA) of 17 systemic
agents for moderate-to-severe plaque psoriasis. It fits **two independent
models** with the same structure, one per end point, and this package ships
each as its own model file:

* `He_2021_psoriasis_pasi75_mbma` -- the proportion of patients reaching a 75%
  reduction from baseline PASI (PASI75), 17 drugs (Table 2 and Supplementary
  Table S1).
* `He_2021_psoriasis_pasi90_mbma` -- the same for a 90% reduction (PASI90),
  16 drugs; alefacept had too few PASI90 data (Table 4 and Supplementary
  Table S2).

The structure follows the methodology of Checchio 2017
(`Checchio_2017_psoriasis_pasi75_longitudinal_mbma`), which He 2021 cites for
it. The logit of the responder rate is a placebo term that rises
exponentially to a plateau plus, for each drug, an Emax dose-response
multiplied by an exponential onset. Body weight acts on the placebo plateau.

```{r models}
mod75 <- readModelDb("He_2021_psoriasis_pasi75_mbma")
mod90 <- readModelDb("He_2021_psoriasis_pasi90_mbma")
ui75 <- rxode2::rxode2(mod75)
ui90 <- rxode2::rxode2(mod90)
```

# Population

Both models operate at the **study-arm** level: one observation is one trial
arm's responder proportion at one time, not one patient's outcome. The random
effect is *between-study*, and neither model may be used to simulate individual
patients.

The dataset is 80 randomised placebo- or active-controlled trials, 235
treatment arms and 40,323 patients, identified from PubMed, Cochrane, Embase and
ClinicalTrials.gov up to 18 July 2019 (Methods 'Database Development'; Results
'Characteristics of Included Studies'). PASI75 was reported in 233 arms and
PASI90 in 224. Arm-level demographics are summarised as medians (range) in
Table 1: 69.05% male, body weight 89.6 kg (66.6-99), age 45 years (38.6-55.3),
baseline PASI 20 (11-33.1).

```{r population}
pop <- mod75()$population
tibble::tibble(
  Field = c("Species", "Trials", "Patients", "Arms (PASI75 / PASI90)",
            "Disease state", "Weight", "Female (%)"),
  Value = c(pop$species, pop$n_studies, pop$n_subjects, "233 / 224",
            pop$disease_state, pop$weight_range, pop$sex_female_pct)
) |>
  knitr::kable()
```

# Source trace

Every model equation and parameter block, with where it came from. The in-file
comments carry the same trace per parameter.

```{r source-trace-equations}
tibble::tribble(
  ~Component, ~Source,
  "N_response ~ binomial(N, P); P = inverse-logit(E0 + Edrug)", "Equations 1-3, Methods 'Model Development'",
  "E0 = BSL + A * (1 - exp(-kpbo * time * exp(eta)))", "Equation 4 (eta inside the exponent; see Assumptions)",
  "Edrug = Emax * (1 - exp(-k * time)) * dose / (dose + ED50)", "Equation 5, Hill coefficient c fixed to 1 per Methods",
  "ED50 = 0 FIX makes dose / (dose + ED50) = 1: written as (dose > 0)", "Results; Table 2 and Table 4 '0 FIX' entries",
  "Obs = P + Weight * eps; Weight = sqrt(P * (1 - P) / N)", "Equations 7-8",
  "A_i = A * (WT / 90)^theta", "Equation 9; 90 kg per the Figure 2-4 captions"
) |>
  knitr::kable(caption = "Structural, covariate and residual equations.")
```

```{r source-trace-params}
tibble::tribble(
  ~Model, ~Parameters, ~Source,
  "PASI75", "BSL, A, kpbo, weight effect on A, omega, sigma", "Supplementary Table S1",
  "PASI75", "Emax, ED50, k for 17 drugs (ED50 0 FIX for risankizumab, alefacept, methotrexate; apremilast Emax 9 FIX)", "Table 2",
  "PASI90", "BSL, A, kpbo, weight effect on A, omega, sigma", "Supplementary Table S2",
  "PASI90", "Emax, ED50, k for 16 drugs (ED50 0 FIX for ustekinumab, methotrexate; apremilast Emax 8.8 FIX)", "Table 4"
) |>
  knitr::kable(caption = "Parameter provenance by block.")
```

The three display equations that carry the model (4, 5 and 9) are typeset as
images in the published PDF and are lost by a plain-text conversion; they were
read from the PDF's text layer, and the position of `exp(eta)` in Equation 4 was
checked from the glyph positions (it sits on the superscript baseline with
`-kpbo`, i.e. inside the exponent).

# Regimens

Both models take dose through one `CONMED_<drug>_DOSE` covariate column per
drug and zero in every other column. The value is the **maintenance dose per
administration** (mg; infliximab mg/kg). The regimens below are the rows of
Tables 3 and 5; the loading doses in their labels do not enter the model.

```{r arms}
doseCols <- c(
  "CONMED_ADALIMUMAB_DOSE", "CONMED_INFLIXIMAB_DOSE", "CONMED_ETANERCEPT_DOSE",
  "CONMED_CERTOLIZUMAB_DOSE", "CONMED_USTEKINUMAB_DOSE", "CONMED_BRIAKINUMAB_DOSE",
  "CONMED_GUSELKUMAB_DOSE", "CONMED_TILDRAKIZUMAB_DOSE", "CONMED_RISANKIZUMAB_DOSE",
  "CONMED_SECUKINUMAB_DOSE", "CONMED_IXEKIZUMAB_DOSE", "CONMED_BRODALUMAB_DOSE",
  "CONMED_APREMILAST_DOSE", "CONMED_TOFACITINIB_DOSE", "CONMED_BARICITINIB_DOSE",
  "CONMED_ALEFACEPT_DOSE", "CONMED_MTX_DOSE"
)

arms <- tibble::tribble(
  ~arm,                    ~regimen,                     ~col,                        ~dose, ~cls,
  "Adalimumab",            "80 mg wk 0, 40 mg q2w",      "CONMED_ADALIMUMAB_DOSE",       40, "TNF-alpha",
  "Infliximab",            "5 mg/kg wk 0, 2, 6, q8w",    "CONMED_INFLIXIMAB_DOSE",        5, "TNF-alpha",
  "Etanercept",            "50 mg biw",                  "CONMED_ETANERCEPT_DOSE",       50, "TNF-alpha",
  "Certolizumab pegol",    "400 mg wk 0, 2, 4, 200 mg q2w", "CONMED_CERTOLIZUMAB_DOSE", 200, "TNF-alpha",
  "Ustekinumab",           "45 mg wk 0, 4, q12w",        "CONMED_USTEKINUMAB_DOSE",      45, "IL-12/23",
  "Briakinumab",           "100 mg q4w",                 "CONMED_BRIAKINUMAB_DOSE",     100, "IL-12/23",
  "Guselkumab",            "100 mg wk 0, 4, q8w",        "CONMED_GUSELKUMAB_DOSE",      100, "IL-23",
  "Tildrakizumab",         "100 mg wk 0, 4, q12w",       "CONMED_TILDRAKIZUMAB_DOSE",   100, "IL-23",
  "Risankizumab",          "150 mg wk 0, 4, q12w",       "CONMED_RISANKIZUMAB_DOSE",    150, "IL-23",
  "Secukinumab",           "300 mg wk 0-4, q4w",         "CONMED_SECUKINUMAB_DOSE",     300, "IL-17",
  "Ixekizumab 160",        "160 mg wk 0, q4w",           "CONMED_IXEKIZUMAB_DOSE",      160, "IL-17",
  "Ixekizumab 80",         "160 mg wk 0, 80 mg q4w",     "CONMED_IXEKIZUMAB_DOSE",       80, "IL-17",
  "Brodalumab",            "210 mg wk 0, 1, 2, q2w",     "CONMED_BRODALUMAB_DOSE",      210, "IL-17",
  "Apremilast",            "30 mg b.i.d.",               "CONMED_APREMILAST_DOSE",       30, "PDE4",
  "Tofacitinib 5",         "5 mg b.i.d.",                "CONMED_TOFACITINIB_DOSE",       5, "JAK",
  "Tofacitinib 10",        "10 mg b.i.d.",               "CONMED_TOFACITINIB_DOSE",      10, "JAK",
  "Baricitinib",           "10 mg qd",                   "CONMED_BARICITINIB_DOSE",      10, "JAK",
  "Alefacept",             "10 mg qw",                   "CONMED_ALEFACEPT_DOSE",        10, "CD2",
  "Methotrexate",          "20 mg qw",                   "CONMED_MTX_DOSE",              20, "DHFR"
)

#' Build an rxode2 input frame: one id per arm, observations at `times`.
#' Only the dose columns the model uses are kept, so the PASI90 frame carries
#' no alefacept column.
makeArmData <- function(armTbl, times, cols, wt = 90, nArm = 100) {
  do.call(rbind, lapply(seq_len(nrow(armTbl)), function(i) {
    d <- data.frame(id = i, time = times, WT = wt, N_ARM = nArm)
    for (cc in cols) d[[cc]] <- 0
    d[[armTbl$col[i]]] <- armTbl$dose[i]
    d
  }))
}

arms90 <- arms |> filter(arm != "Alefacept")
cols90 <- setdiff(doseCols, "CONMED_ALEFACEPT_DOSE")

arms |>
  select(Regimen = arm, `Table 3/5 regimen` = regimen, `Dose column` = col,
         `Dose value` = dose) |>
  knitr::kable(caption = "Regimens of He 2021 Tables 3 and 5 and the maintenance dose each maps to.")
```

# Reproducing Table 3 (PASI75) and Table 5 (PASI90)

Tables 3 and 5 report, for a 90 kg arm at each regimen, the median PASI75 and
PASI90 responder rate over 1,000 simulations at Weeks 4, 8, 12, 16 and 24.
These are **predictions, not parameter estimates**, so reproducing them is an
independent check on the whole encoding: 95 PASI75 cells and 90 PASI90 cells.

```{r typical-sim, message = FALSE, warning = FALSE}
# zeroRe() removes every random effect on purpose, so rxode2's 'no sigma' and
# 'multi-subject simulation without omega' warnings are expected here.
weeks <- c(4, 8, 12, 16, 24)
tGrid <- sort(unique(c(seq(0, 24, by = 0.25), weeks)))
sol75 <- rxode2::rxSolve(rxode2::zeroRe(ui75), makeArmData(arms, tGrid, doseCols),
                         returnType = "data.frame") |>
  mutate(arm = arms$arm[id])
sol90 <- rxode2::rxSolve(rxode2::zeroRe(ui90), makeArmData(arms90, tGrid, cols90),
                         returnType = "data.frame") |>
  mutate(arm = arms90$arm[id])
```

```{r table3}
# He 2021 Table 3, medians (%), transcribed.
table3 <- tibble::tribble(
  ~arm,                  ~w4,   ~w8,   ~w12,  ~w16,  ~w24,
  "Adalimumab",          21.30, 56.95, 67.80, 71.20, 72.80,
  "Infliximab",          34.10, 67.10, 75.65, 78.20, 79.70,
  "Etanercept",           7.98, 34.40, 49.60, 55.00, 57.80,
  "Certolizumab pegol",  15.30, 53.15, 67.70, 71.80, 74.00,
  "Ustekinumab",         10.75, 47.30, 65.30, 71.20, 74.50,
  "Briakinumab",         21.50, 66.80, 79.50, 82.80, 84.40,
  "Guselkumab",          20.65, 67.00, 80.40, 83.90, 85.80,
  "Tildrakizumab",        9.26, 42.80, 61.00, 68.10, 71.20,
  "Risankizumab",        21.15, 72.50, 85.95, 89.40, 91.00,
  "Secukinumab",         39.60, 77.20, 84.50, 86.70, 87.60,
  "Ixekizumab 160",      56.50, 80.40, 85.90, 87.70, 88.50,
  "Ixekizumab 80",       48.95, 75.20, 82.40, 84.60, 85.70,
  "Brodalumab",          48.55, 74.95, 81.70, 84.00, 85.20,
  "Apremilast",           5.68, 20.30, 29.50, 33.20, 35.30,
  "Tofacitinib 5",        9.07, 29.00, 39.20, 42.80, 45.10,
  "Tofacitinib 10",      16.30, 46.55, 57.70, 61.60, 63.10,
  "Baricitinib",          7.34, 33.10, 48.70, 54.80, 58.30,
  "Alefacept",            1.90,  7.46, 12.90, 15.90, 19.30,
  "Methotrexate",         4.14, 19.00, 30.80, 36.80, 40.10
)

# He 2021 Table 5, medians (%), transcribed.
table5 <- tibble::tribble(
  ~arm,                  ~w4,   ~w8,   ~w12,  ~w16,  ~w24,
  "Adalimumab",           6.10, 30.30, 43.50, 48.05, 50.60,
  "Infliximab",          10.40, 36.80, 49.15, 53.40, 55.90,
  "Etanercept",           1.16,  9.90, 22.00, 28.95, 34.40,
  "Certolizumab pegol",   2.29, 20.00, 36.50, 43.50, 47.80,
  "Ustekinumab",          2.64, 22.90, 41.60, 48.85, 53.30,
  "Briakinumab",          5.83, 37.10, 54.85, 60.80, 63.60,
  "Guselkumab",          11.70, 47.45, 61.00, 65.40, 67.40,
  "Tildrakizumab",        2.21, 18.10, 34.80, 40.70, 44.90,
  "Risankizumab",         5.11, 42.15, 65.50, 72.70, 76.70,
  "Secukinumab",         10.60, 47.95, 62.50, 66.70, 69.30,
  "Ixekizumab 160",      21.60, 55.35, 67.20, 71.10, 73.10,
  "Ixekizumab 80",       18.40, 50.40, 62.60, 66.95, 69.40,
  "Brodalumab",          19.90, 52.15, 64.50, 68.60, 70.80,
  "Apremilast",           0.38,  2.68,  6.09,  9.75, 16.40,
  "Tofacitinib 5",        1.99, 12.00, 19.50, 23.40, 25.10,
  "Tofacitinib 10",       3.52, 21.40, 33.60, 38.15, 40.70,
  "Baricitinib",          3.05, 17.20, 26.20, 29.95, 32.30,
  "Methotrexate",         0.85,  6.27, 13.05, 17.00, 20.10
)

#' Long-format comparison of a published table against the typical solve.
compareTable <- function(pub, sol, outcome) {
  pubLong <- pub |>
    pivot_longer(-arm, names_to = "week", values_to = "published") |>
    mutate(week = as.numeric(sub("w", "", week)))
  modLong <- sol |>
    filter(time %in% weeks) |>
    transmute(arm, week = time, model = 100 * .data[[outcome]])
  pubLong |>
    left_join(modLong, by = c("arm", "week")) |>
    mutate(diff = model - published)
}

cmp75 <- compareTable(table3, sol75, "prob_pasi75")
cmp90 <- compareTable(table5, sol90, "prob_pasi90")

cmp75 |>
  filter(week %in% c(4, 12, 24)) |>
  mutate(cell = sprintf("%.1f / %.1f", published, model)) |>
  select(arm, week, cell) |>
  pivot_wider(names_from = week, values_from = cell) |>
  rename(Regimen = arm, `Week 4 (pub / model)` = `4`,
         `Week 12 (pub / model)` = `12`, `Week 24 (pub / model)` = `24`) |>
  knitr::kable(caption = "PASI75 (%) against He 2021 Table 3, 90 kg arm.")

cmp90 |>
  filter(week %in% c(4, 12, 24)) |>
  mutate(cell = sprintf("%.1f / %.1f", published, model)) |>
  select(arm, week, cell) |>
  pivot_wider(names_from = week, values_from = cell) |>
  rename(Regimen = arm, `Week 4 (pub / model)` = `4`,
         `Week 12 (pub / model)` = `12`, `Week 24 (pub / model)` = `24`) |>
  knitr::kable(caption = "PASI90 (%) against He 2021 Table 5, 90 kg arm.")
```

```{r table-gate}
# Deterministic comparison: the model side is a typical-value solve at the
# published parameters, the published side is fixed text. Nothing is drawn at
# random, so a tight bound over every cell is the right gate.
stopifnot(
  nrow(cmp75) == 95, !anyNA(cmp75$model),
  nrow(cmp90) == 90, !anyNA(cmp90$model),
  # PASI75 must be ~0 at randomisation: it is a reduction from each arm's own
  # baseline.
  all(sol75$prob_pasi75[sol75$time == 0] < 1e-3),
  max(abs(cmp75$diff)) < 1.0,
  max(abs(cmp90$diff)) < 1.5
)
cat(sprintf(
  "PASI75: max |model - Table 3| = %.2f pp over %d cells (mean signed %+.2f)\nPASI90: max |model - Table 5| = %.2f pp over %d cells (mean signed %+.2f)\n",
  max(abs(cmp75$diff)), nrow(cmp75), mean(cmp75$diff),
  max(abs(cmp90$diff)), nrow(cmp90), mean(cmp90$diff)
))
```

Every cell is reproduced to within 1 percentage point for PASI75 and 1.5 points
for PASI90. The typical-value solve sits slightly *above* the published
medians on average, more so for PASI90. That is expected rather than a
transcription error: the published values are medians of 1,000 simulations
that carry random effects and, going by the width of the reported intervals,
parameter uncertainty, and those medians need not coincide with the
typical-value prediction. The same selection of maintenance doses reproduces
both tables, which is what settles the dose metric (see Assumptions).

## Figures 2 and 4: time courses at the clinical dose

```{r fig2, fig.cap = "Replicates Figure 2 of He 2021: model-predicted typical PASI75 time course by regimen, 90 kg arm."}
sol75 |>
  left_join(select(arms, arm, cls), by = "arm") |>
  ggplot(aes(time, 100 * prob_pasi75, colour = arm)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~cls) +
  labs(x = "Time (weeks)", y = "PASI75 responders (%)", colour = NULL) +
  coord_cartesian(ylim = c(0, 100)) +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r fig4, fig.cap = "Replicates Figure 4 of He 2021: model-predicted typical PASI90 time course by regimen, 90 kg arm."}
sol90 |>
  left_join(select(arms, arm, cls), by = "arm") |>
  ggplot(aes(time, 100 * prob_pasi90, colour = arm)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~cls) +
  labs(x = "Time (weeks)", y = "PASI90 responders (%)", colour = NULL) +
  coord_cartesian(ylim = c(0, 100)) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Figure 3: Week-12 ranking

```{r fig3, fig.cap = "Replicates the median points of Figure 3 of He 2021: Week-12 PASI75 (A) and PASI90 (B) by regimen, typical 90 kg arm, ranked from high to low."}
rank12 <- bind_rows(
  cmp75 |> filter(week == 12) |> mutate(endpoint = "A: PASI75"),
  cmp90 |> filter(week == 12) |> mutate(endpoint = "B: PASI90")
)
rank12 |>
  mutate(arm = reorder(paste(arm, endpoint, sep = "___"), model)) |>
  ggplot(aes(model, arm)) +
  geom_point(colour = "black") +
  geom_point(aes(x = published), shape = 1, colour = "red", size = 2.5) +
  scale_y_discrete(labels = function(x) sub("___.*$", "", x)) +
  facet_wrap(~endpoint, scales = "free_y") +
  labs(x = "Week-12 responders (%): model (filled) and Table 3 / 5 (open)", y = NULL) +
  theme_bw()

top75 <- rank12 |> filter(endpoint == "A: PASI75") |> arrange(desc(model))
top90 <- rank12 |> filter(endpoint == "B: PASI90") |> arrange(desc(model))
stopifnot(
  # Results: risankizumab then ixekizumab 160 mg lead PASI75 at Week 12, and
  # ixekizumab 160 mg then risankizumab lead PASI90.
  identical(top75$arm[1:2], c("Risankizumab", "Ixekizumab 160")),
  identical(top90$arm[1:2], c("Ixekizumab 160", "Risankizumab")),
  # Alefacept is the least efficacious PASI75 regimen and apremilast the least
  # efficacious PASI90 regimen.
  tail(top75$arm, 1) == "Alefacept",
  tail(top90$arm, 1) == "Apremilast"
)
```

The ordering statements in the Results -- risankizumab then ixekizumab 160 mg
at the top of the Week-12 PASI75 ranking, the reverse for PASI90, alefacept and
apremilast at the bottom -- all hold for the packaged models.

# Where the between-study effect acts

Equation 4 prints the between-study random effect inside the exponent, on the
placebo onset rate: `exp(-kpbo * time * exp(eta))`. Supplementary Tables S1
and S2 label the same variance row `omega(A), %`, which would put it on the
asymptote A. The two readings give identical typical values, so Tables 3 and 5
cannot separate them. The paper's own visual predictive check (Supplementary
Figure S3) can. That figure is a vector graphic, so its percentile lines were
read exactly from the PDF's drawing coordinates. For the risankizumab panel
(one regimen, ED50 fixed to 0, so no dose mixing) they give:

```{r vpc-digitised}
vpcRis <- tibble::tribble(
  ~week, ~p025, ~p50, ~p975,
  4,       7.6, 21.5,  43.0,
  22,     84.0, 91.3,  99.1
)
knitr::kable(vpcRis, caption = "Risankizumab PASI75 VPC percentiles (%), Supplementary Figure S3, read from the vector drawing.")
```

At Week 22 the placebo term has reached its plateau, so a random effect on the
onset rate barely moves the prediction, while one on the asymptote moves it in
full. The deterministic calculation below evaluates the typical risankizumab
arm with the random effect held at its 2.5th and 97.5th percentiles
(`+/- 1.96 omega`) under each reading, with no residual error.

```{r eta-placement}
p75 <- setNames(ui75$iniDf$est, ui75$iniDf$name)
omega <- sqrt(p75[["eta_study_lkpbo"]])

#' Typical risankizumab PASI75 (%) at `t` weeks with the random effect `eta`
#' on the onset rate (`where = "kpbo"`) or on the asymptote (`where = "A"`).
risPasi75 <- function(t, eta, where) {
  kpbo <- exp(p75[["lkpbo"]] + if (where == "kpbo") eta else 0)
  asym <- p75[["asym_pbo"]] * if (where == "A") exp(eta) else 1
  e0 <- p75[["bsl_pbo"]] + asym * (1 - exp(-kpbo * t))
  ed <- p75[["emax_risankizumab"]] * (1 - exp(-exp(p75[["lkdrug_risankizumab"]]) * t))
  100 * stats::plogis(e0 + ed)
}

placement <- expand.grid(week = c(4, 22), where = c("kpbo", "A"),
                         stringsAsFactors = FALSE) |>
  rowwise() |>
  mutate(
    lo = risPasi75(week, -1.96 * omega, where),
    typical = risPasi75(week, 0, where),
    hi = risPasi75(week, 1.96 * omega, where)
  ) |>
  ungroup()

placement |>
  mutate(where = ifelse(where == "kpbo", "onset rate kpbo (Equation 4)", "asymptote A (table label)")) |>
  rename(Week = week, `Random effect on` = where, `2.5th (%)` = lo,
         `Typical (%)` = typical, `97.5th (%)` = hi) |>
  knitr::kable(digits = 1, caption = "Spread of the risankizumab arm from the between-study effect alone.")

wk22 <- placement |> filter(week == 22)
stopifnot(
  # On the asymptote, the effect alone pushes the Week-22 2.5th percentile far
  # below the VPC's 84.0%, before any residual error widens it further.
  wk22$lo[wk22$where == "A"] < 70,
  # On the onset rate, the effect has all but vanished by Week 22 (about one
  # percentage point either way).
  wk22$typical[wk22$where == "kpbo"] - wk22$lo[wk22$where == "kpbo"] < 2
)

# Arm size at which the residual alone (sigma = 1.33 times the binomial SE)
# places the 2.5th percentile at the VPC's 84.0%.
pTyp <- wk22$typical[wk22$where == "kpbo"] / 100
nImplied <- (1.96 * p75[["addSd_prob_pasi75"]] * sqrt(pTyp * (1 - pTyp)) / (pTyp - 0.840))^2
cat(sprintf("Residual-only reading of the Week-22 VPC band implies an arm size of about %.0f patients.\n", nImplied))
```

Under the table's label the between-study effect alone would put the Week-22
2.5th percentile at `r round(wk22$lo[wk22$where == "A"])`%, far outside the
published band, and residual error could only widen it. Under Equation 4 the
effect moves Week 22 by barely a percentage point and the band is residual error, which reproduces the
VPC's 84.0% lower line for arms of about `r round(nImplied, -1)` patients, a
plausible arm size for the phase 2 and 3 risankizumab trials pooled in that
panel.
Both models therefore place the random effect as Equation 4 prints it.

# Between-study variability

Simulating study arms rather than typical values shows the spread a trial
designer should expect. The cohort is 150 risankizumab arms of 100 patients
each (within the 200-per-arm cap), with the between-study effect and the
residual error drawn; the lines of the Supplementary Figure S3 panel are
overlaid.

```{r bsv, fig.cap = "Simulated spread of 150 risankizumab 150 mg arms of 100 patients (grey: 2.5th-97.5th percentile of the simulated observations; black: median). Red points: the Supplementary Figure S3 VPC percentiles."}
rxode2::rxSetSeed(20210701)
nArms <- 150
bsvArms <- arms |> filter(arm == "Risankizumab") |> slice(rep(1, nArms))
bsvSol <- rxode2::rxSolve(ui75, makeArmData(bsvArms, 0:24, doseCols, nArm = 100),
                          returnType = "data.frame")

bsvQ <- bsvSol |>
  group_by(time) |>
  summarise(lo = stats::quantile(sim, 0.025), md = stats::median(sim),
            hi = stats::quantile(sim, 0.975), .groups = "drop")

ggplot(bsvQ, aes(time)) +
  geom_ribbon(aes(ymin = 100 * lo, ymax = 100 * hi), fill = "grey80") +
  geom_line(aes(y = 100 * md), linewidth = 0.8) +
  geom_point(data = pivot_longer(vpcRis, -week), aes(week, value), colour = "red") +
  labs(x = "Time (weeks)", y = "PASI75 responders (%)") +
  coord_cartesian(ylim = c(0, 100)) +
  theme_bw()

typRis <- sol75 |> filter(arm == "Risankizumab", time == 12)
stopifnot(
  # Central tendency only: the simulated extremes are not reproducible across
  # rxode2 builds.
  abs(100 * (bsvQ$md[bsvQ$time == 12] - typRis$prob_pasi75)) < 3,
  abs(100 * bsvQ$md[bsvQ$time == 22] - vpcRis$p50[vpcRis$week == 22]) < 3
)
```

# Body weight

Body weight enters only the placebo asymptote, through a negative power
(PASI75 -0.245, PASI90 -0.214), so heavier arms have a lower placebo plateau
and, because the drug term adds on the same logit scale, a slightly lower
total response.

```{r weight, message = FALSE, warning = FALSE}
# Typical-value solve; see the typical-sim chunk for the expected warnings.
wtDat <- bind_rows(lapply(c(70, 90, 110), function(w) {
  makeArmData(arms |> filter(arm == "Adalimumab"), 24, doseCols, wt = w) |>
    mutate(id = w)
}))
wtSol <- rxode2::rxSolve(rxode2::zeroRe(ui75), wtDat, returnType = "data.frame")
wtSol |>
  transmute(`Arm mean weight (kg)` = id, `Adalimumab Week-24 PASI75 (%)` = 100 * prob_pasi75) |>
  knitr::kable(digits = 1)
stopifnot(all(diff(wtSol$prob_pasi75) < 0))
```

# NCA

These models have no pharmacokinetic layer: dose enters as a covariate, there
are no dose events and no concentrations, so a non-compartmental analysis does
not apply. The validation above compares the models against every predicted
value the paper publishes instead.

# Assumptions and deviations

## Dose is the per-administration maintenance dose

Tables 1, 3 and 5 give regimens, not the number the model reads. The dose
metric is settled by the published predictions: ixekizumab 160 mg and 80 mg
maintenance share one parameter set and are reproduced only by Dose = 160 and
80; tofacitinib 5 and 10 mg b.i.d. only by Dose = 5 and 10 (not the daily
totals); apremilast 30 mg b.i.d. by Dose = 30 (Dose = 60 gives a Week-12 PASI75
of 61% against a published 29.5%); adalimumab by the 40 mg maintenance dose,
not the 80 mg loading dose (83% against 67.8%). The Results text says the same
for adalimumab ("the clinical dosage is 40 mg every 2 weeks"). Certolizumab
pegol's '400 mg 0, 2, 4, 200 mg q2w' row is closer with Dose = 200 (68.0%
against 67.7%) than with 400 (70.0%). Infliximab is mg/kg.

```{r dose-metric, message = FALSE, warning = FALSE}
# Typical-value solves; see the typical-sim chunk for the expected warnings.
alt <- tibble::tribble(
  ~arm,                 ~col,                       ~chosen, ~alternative, ~published,
  "Adalimumab",         "CONMED_ADALIMUMAB_DOSE",        40,           80,      67.80,
  "Certolizumab pegol", "CONMED_CERTOLIZUMAB_DOSE",     200,          400,      67.70,
  "Apremilast",         "CONMED_APREMILAST_DOSE",        30,           60,      29.50,
  "Tofacitinib 10",     "CONMED_TOFACITINIB_DOSE",       10,           20,      57.70
)
#' Week-12 typical PASI75 (%) for one drug column at one dose.
wk12At <- function(col, dose) {
  d <- makeArmData(tibble::tibble(col = col, dose = dose), 12, doseCols)
  100 * rxode2::rxSolve(rxode2::zeroRe(ui75), d, returnType = "data.frame")$prob_pasi75
}
alt <- alt |>
  rowwise() |>
  mutate(modelChosen = wk12At(col, chosen), modelAlternative = wk12At(col, alternative)) |>
  ungroup()
alt |>
  select(Regimen = arm, `Table 3 Week 12 (%)` = published,
         `Dose used` = chosen, `Model (%)` = modelChosen,
         `Alternative dose` = alternative, `Model at alternative (%)` = modelAlternative) |>
  knitr::kable(digits = 1)
stopifnot(all(abs(alt$modelChosen - alt$published) < abs(alt$modelAlternative - alt$published)))
```

## Where the between-study random effect acts

Equation 4 places `exp(eta)` on the placebo onset rate; Supplementary Tables
S1 and S2 label the variance `omega(A)`. The equation is followed, because the
paper's own VPC is incompatible with the table label (see "Where the
between-study effect acts"). Typical-value predictions are unaffected either
way.

## Random-effect and residual scale

The random effect is printed as a CV (24.9% for PASI75, 26% for PASI90) and is
converted to a log-normal variance by `log(1 + CV^2)` (0.0602 and 0.0654).
Reading the CV instead as `sqrt(omega^2)` gives 0.0620 and 0.0676, an
immaterial difference. The residual term is printed as `sigma` = 1.33 in both
tables and is read as a standard deviation, the symbol the Methods use for the
square root of the residual variance; the variance reading would give 1.153.
Either way it multiplies the binomial standard error `sqrt(P(1 - P)/N)`, so
simulating it needs the arm size in an `N_ARM` column.

## Body weight centring

Equation 9 divides by "mean(covariate)" without a number. 90 kg is used: the
captions of Figures 2-4 call 90 kg the average body weight, all published
predictions are made at 90 kg, and the Table 1 dataset median is 89.6 kg
(which would change the weight factor by 0.1%). Arms that did not report
weight were set to the dataset median by the authors; downstream users should
do the same.

## ED50 fixed at zero

Where the authors fixed an ED50 to 0 (PASI75: risankizumab, alefacept,
methotrexate; PASI90: ustekinumab, methotrexate), `dose / (dose + 0)` is 1 for
every positive dose, so the model reads only whether the dose column is
positive. No log-scale ED50 parameter is declared for these drugs because a
log scale cannot hold 0. For those drugs the model predicts the response at the
doses studied and nothing about other doses. Risankizumab and ustekinumab
differ between the two models in this respect, exactly as in the paper.

## Arm-level scope

Both models describe study-arm responder proportions. The random effect is
between-study and the residual is the sampling error of an arm proportion;
neither describes individual patients.

## Hill coefficient

Equation 5 carries a Hill coefficient `c`, which the Methods fix to 1 for every
drug. It is therefore not a model parameter.
