---
title: "Oral paclitaxel (He 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Oral paclitaxel (He 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("He_2022_paclitaxel"))$reference`
* Description: `r rxode2::rxode(readModelDb("He_2022_paclitaxel"))$description`
* Article: <https://doi.org/10.1002/psp4.12799> (open access)
* Supplement (Appendix S1: Tables S1-S6 including the final NONMEM control
  stream; Appendix S2: the analysis dataset):
  <https://europepmc.org/article/PMC/PMC9286714>

Oraxol is oral paclitaxel given with encequidar, a minimally absorbed
P-glycoprotein inhibitor that lets paclitaxel be absorbed from the gut. He
2022 pooled seven studies of oral paclitaxel with encequidar in patients with
solid tumors. The final model has two compartments, first-order absorption
after a short fixed lag, and linear elimination. Bioavailability is fixed at
0.119 for the capsule, so the clearance and volumes are systemic values given
that F. Two covariates were retained: the oral solution's higher
bioavailability, and a larger central volume in non-Asian patients.

# Population

The model was developed from 197 adults with advanced or metastatic solid
tumors (gastric cancer, metastatic breast cancer, lung cancer and others)
enrolled in seven studies in the Republic of Korea, the United States, New
Zealand, Australia and Taiwan (He 2022 Table S1). They contributed 4322
plasma paclitaxel concentrations. Mean age was 59.6 years (range 32-81),
mean weight 67.2 kg (38-139) and mean body surface area 1.73 m^2
(1.29-2.46). 52.3% were female; 55.8% were Asian and 41.1% Caucasian. 59.9%
had mild or moderate renal impairment and 10.2% had (mostly mild) hepatic
impairment (He 2022 Table 1). Oral doses ranged from 60 to 420 mg/m^2. The
two bioavailability studies, and the phase III regimen, used 205 mg/m^2 once
daily on days 1-3 of each week. 92.4% of patients took capsules; the other
15, all in the first study, took the intravenous paclitaxel solution by
mouth.

The same information is available programmatically via
`readModelDb("He_2022_paclitaxel")()$population`.

# Source trace

Every `ini()` value carries an in-file comment pointing to its source in
`inst/modeldb/specificDrugs/He_2022_paclitaxel.R`. The table below collects
them.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL) | log(33.7) L/h | Table 2; Table S4 |
| `lvc` (V2, Asian) | log(50.7) L | Table 2; Table S4 |
| `lq` (Q) | log(40.6) L/h | Table 2; Table S4 |
| `lvp` (V3) | log(855) L | Table 2; Table S4 |
| `lka` (KA) | log(0.724) 1/h | Table 2; Table S4 |
| `ltlag` (ALAG1) | fixed(log(0.215)) h | Table 2 (fixed); Table S6 THETA(7) |
| `lfdepot` (F1, capsule) | fixed(log(0.119)) | Table 2 (fixed); Table S6 THETA(6) |
| `e_race_nonasian_vc` | 0.696 | Table 2 'Race on V2 (proportional)'; Table S6 THETA(8) |
| `e_form_solution_fdepot` | 0.895 | Table 2 'Formulation on F1 (proportional)'; Table S6 THETA(9) |
| `etalcl` | 0.1118 | Table 2 'ETA CL' 34.4 CV%, as log(CV^2 + 1) |
| `etalvc` | 1.4104 | Table 2 'ETA V2' 176 CV%, as log(CV^2 + 1) |
| `etalfdepot` | 0.2073 | Table S4 'BSV on F1' 48.0 CV%, as log(CV^2 + 1); Table S6 eta6 on F1 |
| `expSd` | sqrt(0.208) = 0.4561 | Table 2 'Log additive' 0.208 (SIGMA variance); Table S6 `Y = LOG(F) + ERR(1)` |
| `vc <- ... * (1 + e * (1 - RACE_ASIAN))` | n/a | Table S6 `TVV2 = THETA(2)*(1 + THETA(8)*RACB)`, RACB = non-Asian |
| `fdepot <- ... * (1 + e * FORM_SOLUTION)` | n/a | Table S6 `TVF1 = THETA(6)*(1 + THETA(9)*(2-FORM))`, FORM 1 = solution |
| ODEs, `alag(depot)`, `f(depot)` | n/a | Table S6 `ADVAN4 TRANS4`, Figure 1 |
| `Cc <- central / vc * 1000` | n/a | Table S6 `S2 = V2/1000` (dose in mg, concentration in ng/mL) |

```{r load-model}
mod <- readModelDb("He_2022_paclitaxel")
mod_typical <- mod |> rxode2::zeroRe()
```

# Typical-value checks against printed values

The Results section prints two derived typical values: the non-Asian central
volume, 50.7 x (1 + 0.696) = 86.0 L, and the oral-solution bioavailability,
0.226. Both come out of the model's own `model()` block.

```{r typical-derived}
ev_one <- rxode2::et(amt = 100, cmt = "depot") |>
  rxode2::et(c(0, 1, 2))
derived <- dplyr::bind_rows(
  lapply(
    list(
      c(RACE_ASIAN = 1, FORM_SOLUTION = 0),
      c(RACE_ASIAN = 0, FORM_SOLUTION = 0),
      c(RACE_ASIAN = 1, FORM_SOLUTION = 1)
    ),
    function(cv) {
      s <- rxode2::rxSolve(mod_typical, ev_one, params = cv)
      data.frame(
        RACE_ASIAN = cv[["RACE_ASIAN"]],
        FORM_SOLUTION = cv[["FORM_SOLUTION"]],
        vc = s$vc[1],
        fdepot = s$fdepot[1]
      )
    }
  )
)
knitr::kable(derived, digits = 4, caption = "Typical V2 (L) and F1 by covariate level.")

stopifnot(
  # Printed '50.7 x (1 + 0.696) = 86.0 L' for non-Asian patients.
  abs(derived$vc[2] - 86.0) < 0.05,
  abs(derived$vc[1] - 50.7) < 1e-8,
  # Printed 'bioavailability estimated to be 0.226' for the oral solution.
  abs(derived$fdepot[3] - 0.226) < 0.0005,
  abs(derived$fdepot[1] - 0.119) < 1e-8
)
```

Because the model is linear, the total AUC after a capsule dose is
F x Dose / CL exactly. A typical-value solve recovers that identity. The
Introduction also states that 205 mg/m^2 by mouth on three consecutive days
gives a systemic AUC similar to 80 mg/m^2 intravenous paclitaxel. Under this
model the ratio is 3 x 0.119 x 205 / 80 = 0.915, whatever the clearance. The
intravenous dose is given straight into `central`; F1 applies to `depot`
only.

```{r typical-auc}
bsa_ref <- 1.73
# Fine grid over the fast distribution phase (alpha ~ 1.5 1/h), coarse tail.
auc_grid <- sort(unique(c(seq(0, 200, by = 0.02), seq(200, 3000, by = 0.5))))
ev_oral <- rxode2::et(
  amt = 205 * bsa_ref, cmt = "depot", time = c(0, 24, 48)
) |>
  rxode2::et(auc_grid)
ev_iv <- rxode2::et(
  amt = 80 * bsa_ref, cmt = "central", time = 0, dur = 1
) |>
  rxode2::et(auc_grid)
cov_ref <- c(RACE_ASIAN = 1, FORM_SOLUTION = 0)
s_oral <- as.data.frame(rxode2::rxSolve(
  mod_typical, ev_oral,
  params = cov_ref, rtol = 1e-10, atol = 1e-12
))
s_iv <- as.data.frame(rxode2::rxSolve(
  mod_typical, ev_iv,
  params = cov_ref, rtol = 1e-10, atol = 1e-12
))

trap <- function(t, y) sum(diff(t) * (head(y, -1) + tail(y, -1)) / 2)
auc_oral <- trap(s_oral$time, s_oral$Cc)
auc_iv <- trap(s_iv$time, s_iv$Cc)
auc_oral_cf <- 3 * 0.119 * 205 * bsa_ref / 33.7 * 1000
auc_iv_cf <- 80 * bsa_ref / 33.7 * 1000
data.frame(
  regimen = c("Oral 205 mg/m^2 QD x 3 (capsule)", "IV 80 mg/m^2, 1-h infusion"),
  auc_solver = c(auc_oral, auc_iv),
  auc_closed_form = c(auc_oral_cf, auc_iv_cf)
) |>
  dplyr::rename(
    "Regimen (BSA 1.73 m^2)" = regimen,
    "AUC0-3000h, solver (ng*h/mL)" = auc_solver,
    "F x Dose / CL (ng*h/mL)" = auc_closed_form
  ) |>
  knitr::kable(digits = 0)

stopifnot(
  # Linear trapezoid on the 0.02 h grid: the error scales with the grid
  # spacing squared (0.4% on a 0.5 h grid, so < 1e-4 here); the 3000 h
  # window leaves < 1e-6 of the AUC unsampled.
  abs(auc_oral / auc_oral_cf - 1) < 1e-3,
  abs(auc_iv / auc_iv_cf - 1) < 1e-3,
  # Introduction: oral 205 mg/m^2 x 3 days 'similar to' 80 mg/m^2 IV.
  abs(auc_oral / auc_iv - 3 * 0.119 * 205 / 80) < 1e-3
)
```

# Virtual cohort

He 2022 simulated 1000 capsule patients (500 Asian, 500 non-Asian) on the
phase III regimen: 205 mg/m^2 once daily on days 1-3 of each week for three
weeks. The cohort below has 200 patients, half of them Asian. Body surface
area is drawn from a normal distribution with the Table 1 mean and SD, then
truncated to the Table 1 range. The dose is 205 mg/m^2 times each patient's
body surface area.

```{r cohort}
rxode2::rxSetSeed(20220714)
set.seed(20220714)
n_sub <- 200L
dose_times <- c(0, 24, 48, 168, 192, 216, 336, 360, 384)
obs_times <- seq(0, 504, by = 0.5)

subjects <- tibble::tibble(
  id = seq_len(n_sub),
  RACE_ASIAN = rep(c(1L, 0L), length.out = n_sub),
  FORM_SOLUTION = 0L,
  BSA = pmin(pmax(rnorm(n_sub, 1.73, 0.23), 1.29), 2.46)
) |>
  dplyr::mutate(race = ifelse(RACE_ASIAN == 1, "Asian", "Non-Asian"))

events <- dplyr::bind_rows(
  subjects |>
    tidyr::crossing(time = dose_times) |>
    dplyr::mutate(evid = 1L, amt = 205 * BSA, cmt = "depot"),
  subjects |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(evid = 0L, amt = 0, cmt = "central")
) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  dplyr::mutate(treatment = "Oral 205 mg/m^2 QD x 3 weekly")

stopifnot(
  length(unique(events$id)) == n_sub,
  sum(events$evid == 1) == n_sub * length(dose_times)
)
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep = c("treatment", "race", "BSA")
) |>
  as.data.frame()
# Paclitaxel 0.05 umol/L, the T > 0.05 uM threshold of Table S5
# (molar mass 853.9 g/mol); the paper's figures round it to 40 ng/mL.
c_thr <- 0.05 * 853.9
```

# Replicate published figures

```{r figure-4a, fig.width = 7, fig.height = 4.5}
# Replicates Figure 4a of He 2022: individual simulated concentrations and
# mean + SD on the paper's sampling schedule (pre-dose, 1-4 h after each dose,
# then daily), with the 40 and 10 ng/mL reference lines. Values below the
# 2.5 ng/mL LLOQ are set to missing, as in the paper.
wk <- c(0, 1, 2, 3, 4, 24, 25, 26, 27, 28, 48, 49, 50, 51, 52, 72, 96, 120, 144)
fig_times <- sort(unique(c(wk, wk + 168, wk + 336, 504)))
fig_dat <- sim |>
  dplyr::filter(time %in% fig_times) |>
  dplyr::mutate(Cc = ifelse(Cc < 2.5, NA_real_, Cc))
fig_mean <- fig_dat |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    mean = mean(Cc, na.rm = TRUE),
    sd = sd(Cc, na.rm = TRUE),
    .groups = "drop"
  ) |>
  # Pre-dose at time 0 every value is below the LLOQ, so there is no mean.
  dplyr::filter(is.finite(mean))
ggplot(fig_dat, aes(time, Cc)) +
  geom_point(size = 0.4, alpha = 0.3, na.rm = TRUE) +
  geom_line(data = fig_mean, aes(time, mean), colour = "red") +
  geom_errorbar(
    data = fig_mean,
    aes(x = time, ymin = mean, ymax = mean + sd),
    inherit.aes = FALSE, colour = "red", width = 4
  ) +
  geom_hline(yintercept = c(40, 10), colour = "blue", linetype = "dashed") +
  scale_y_log10() +
  scale_x_continuous(breaks = seq(0, 504, by = 168)) +
  labs(
    x = "Time (h)", y = "Paclitaxel concentration (ng/mL)",
    title = "Oral paclitaxel 205 mg/m^2 QD x 3 for 3 weeks",
    caption = "Replicates Figure 4a of He 2022."
  )
```

The simulated profile has the shape of Figure 4a. Mean concentrations peak
at about 180-200 ng/mL one hour after each dose and fall to a mean of about
15-25 ng/mL by the next dose. Between dosing blocks they decline to a mean
of about 7 ng/mL (quantifiable values only) before the next week's first
dose. The mean profile repeats from week to week with no visible
accumulation, which matches the paper's statement that exposure was
reproducible over weeks 1-4 (Figure S2).

```{r figure-race, fig.width = 7, fig.height = 4}
# Typical profile by race (the only covariate on disposition): the larger
# non-Asian V2 lowers the peak and leaves the AUC unchanged.
ev_wk1 <- rxode2::et(amt = 205 * 1.73, cmt = "depot", time = c(0, 24, 48)) |>
  rxode2::et(seq(0, 168, by = 0.25))
typ <- dplyr::bind_rows(
  as.data.frame(rxode2::rxSolve(mod_typical, ev_wk1,
    params = c(RACE_ASIAN = 1, FORM_SOLUTION = 0)
  )) |> dplyr::mutate(race = "Asian"),
  as.data.frame(rxode2::rxSolve(mod_typical, ev_wk1,
    params = c(RACE_ASIAN = 0, FORM_SOLUTION = 0)
  )) |> dplyr::mutate(race = "Non-Asian")
) |>
  # Drop the pre-absorption rows (Cc = 0 before the lag) for the log axis.
  dplyr::filter(time >= 0.25)
ggplot(typ, aes(time, Cc, colour = race)) +
  geom_line() +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Paclitaxel concentration (ng/mL)", colour = NULL,
    title = "Typical week-1 profile, 205 mg/m^2 QD x 3, BSA 1.73 m^2",
    caption = "Illustrates the race effect on V2 discussed with Figure 2a of He 2022."
  )
```

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)
dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0,
  end = 504,
  auclast = TRUE,
  cmax = TRUE,
  time_above = TRUE,
  # PKNCA reads the time_above threshold from this interval column.
  conc_above = c_thr
)
nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res <- PKNCA::pk.nca(nca_data)
summary(nca_res)
```

## Comparison against published NCA

Table S5 reports NCA of the paper's own simulation of this regimen. The
comparison uses its medians.

```{r nca-compare}
published <- tibble::tribble(
  ~treatment, ~auclast, ~cmax, ~time_above,
  "Oral 205 mg/m^2 QD x 3 weekly", 14168, 268, 93
)
# ncaParamLabel() has no friendly label for 'time_above' and warns; muffle
# only that warning, so any other one still surfaces.
cmp <- withCallingHandlers(
  nlmixr2lib::ncaComparisonTable(
    simulated = nca_res,
    reference = published,
    by = "treatment",
    units = c(auclast = "ng*h/mL", cmax = "ng/mL", time_above = "h"),
    tolerance_pct = 20
  ),
  warning = function(w) {
    if (grepl("time_above", conditionMessage(w))) {
      invokeRestart("muffleWarning")
    }
  }
)
knitr::kable(
  cmp,
  caption = paste(
    "Simulated vs. published (Table S5 median) NCA over 0-504 h.",
    "* differs from reference by >20%."
  )
)
```

The simulated medians come out below Table S5: AUC0-504 by about a sixth,
Cmax by about a third, and time above 0.05 umol/L by about half. This gap does
not point to a transcription error. The maintainers evaluated the published
model on the paper's own analysis dataset (Appendix S2), and its objective
function matches Table 2 (see Assumptions and deviations below). Instead,
the Table S5 numbers depend on how the paper's simulation was done, which
Methods describes only partly:

* NCA was run in Phoenix WinNonlin on concentrations simulated at the sparse
  clinical schedule (pre-dose, 1-4 h after each dose, then every 24 h), with
  values below 2.5 ng/mL set to missing (Figure 4 legend). Trapezoids across
  the 4-24 h gaps after each peak overstate the AUC. On that schedule, this
  model's median AUC0-504 rises to roughly 16,000-18,000 ng*h/mL, depending
  on the trapezoid rule, which brackets the published 14,168.
* The published Cmax and T > 0.05 uM likely include residual error and use
  that sparse grid. The CV of the Table S5 AUC (37%) is also smaller than
  the 34.4% CL and 48.0% F1 variabilities combine to give (about 60%). This
  suggests the paper's simulation did not carry the full IIV structure
  described in Table S4. One possibility is the Table 2 row that labels the
  48.0% IIV as "ETA Q".
* N in Table S5 (264 oral, 135 intravenous) matches neither the stated 1000
  simulated patients nor this cohort. The covariates and doses used are not
  given.

The intravenous arm of Table S5 (175 mg/m^2 over 3 h, mean AUC0-504
16,502 ng*h/mL) cannot come from this model either. With the linear
CL = 33.7 L/h, a 175 mg/m^2 dose at 1.73 m^2 gives AUC = Dose / CL =
`r round(175 * 1.73 / 33.7 * 1000)` ng*h/mL. That is about half the
published value, in line with the saturable clearance of intravenous
paclitaxel at that dose, which the oral model does not describe. The
paper does not say which model it used for the intravenous simulation, so
that arm is not reproduced here.

For these reasons the vignette gates on the model-level identities above,
not on Table S5.

# Assumptions and deviations

* **IIV on F1, not on Q.** Table 2 labels the third IIV row "ETA Q (CV%)
  48.0". Table S4 calls the same row "BSV on F1", and the final control
  stream (Table S6) estimates eta6 on F1 with the eta on Q fixed to 0. The
  model places this IIV on F1. Evaluating the objective function on the
  Appendix S2 data confirms it: OFV -1278.7 with the IIV on F1, against
  -926.8 with it on Q.
* **Variance scales.** Table 2 gives the IIV as CV% and the residual error as
  "Log additive 0.208". The maintainers evaluated the final model, with
  parameters held at the published values, on the Appendix S2 analysis
  dataset restricted to the 197 patients and 4322 observations of the
  analysis (rollover-patient records and the eighth, later study excluded;
  BLQ records dropped as in Table S6). The four readings of the scales give
  these objective function values:

  | IIV as | Residual 0.208 as | OFV |
  |--------|-------------------|-----|
  | log(CV^2 + 1) | variance (SD 0.456) | -1278.7 |
  | (CV/100)^2 | variance (SD 0.456) | -1246.6 |
  | log(CV^2 + 1) | SD | 7206.8 |
  | (CV/100)^2 | SD | 7218.0 |

  Only the first reading matches the printed OFV of 1279.4 to within 1 unit
  in magnitude. The model therefore uses omega^2 = log(CV^2 + 1) and a
  log-scale residual SD of sqrt(0.208). Table 2 prints the OFV without a
  minus sign. For log-transformed data with a log-scale residual variance
  of 0.208 the OFV is negative, so the printed value is taken to be missing
  its sign. The dataset is not redistributed with the package. Its licence
  (CC BY-NC-ND) does not allow it, so this check is not re-run here.
* **Race coding.** The control stream's `RACB` is 1 for every non-Asian race
  code (Caucasian, African American, American Indian, Native Hawaiian or
  Other Pacific Islander) and 0 for Asian. The model uses `1 - RACE_ASIAN`
  in its place. The Appendix S2 dataset confirms that `RACE = 0` is Asian
  (110 patients, matching Table 1).
* **Formulation.** `FORM_SOLUTION = 2 - FORM`, with source `FORM` 1 =
  solution and 2 = capsule (Figure 3 legend). Use `FORM_SOLUTION = 0` for
  the marketed capsule.
* **Encequidar.** Every observation in the analysis came from paclitaxel
  given with encequidar. Encequidar dose and timing were screened on F1
  (Table S3) and not retained. The model does not apply to oral paclitaxel
  without encequidar.
* **Food.** 97% of patients were dosed fasted, and the paper draws no
  conclusion on food. The model describes fasted dosing.
* **Virtual cohort.** Body surface area was drawn from a truncated normal
  distribution using the Table 1 mean, SD and range. Doses were not rounded
  to capsule strengths.
* **Errata.** No correction notice for this article was found in Europe PMC
  as of 2026-10-02.
