---
title: "Gentamicin (He 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gentamicin (He 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("He_2022_gentamicin"))$reference`
* Description: `r rxode2::rxode(readModelDb("He_2022_gentamicin"))$description`
* Article: [Drug Des Devel Ther. 2022;16:13-22](https://doi.org/10.2147/DDDT.S343385)

# Population

He 2022 did not collect new data. A PubMed search (to 14 May 2021) found four
gentamicin PK studies in patients on continuous renal replacement therapy
(CRRT); two reported individual concentrations together with complete dosing
and CRRT settings and were pooled: Petejova et al (7 patients on continuous
venovenous hemofiltration, CVVH) and D'Arcy et al (7 patients on continuous
venovenous hemodiafiltration, CVVHDF). The 14 patients contributed 151
gentamicin concentrations. All were critically ill with acute kidney injury and
septic shock; 9 were male, median age 68.5 years (range 34-79), median weight
72 kg (range 48-102), median APACHE II 28.5 (Table 1).

Gentamicin was given q24h as a 30-minute infusion: 2.4-3.3 mg/kg in the
Petejova patients and 5 mg/kg in the D'Arcy patients. Blood flow was
200 mL/min throughout; the CRRT dose ranged 36.6-67.7 mL/kg/h (median 45) and
the calculated CRRT clearance 1.6-3.5 L/h (median 2.6).

The same information is available programmatically via the model's
`population` metadata (`readModelDb("He_2022_gentamicin")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/He_2022_gentamicin.R`. The table
below collects them in one place.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `CL_total = CL_body * exp(eta) + CL_CRRT` | n/a | Equation 1; Methods, "Population PK Model Development" |
| `CL_CRRT` from CRRT settings (`QE * Sc` or `QE * Sd`, / 1000) | n/a | Equations 2-8 |
| One-compartment, linear elimination | n/a | Results, "Population PK" |
| `lcl` (CL_body) | 1.20 L/h (RSE 20.3%) | Table 2 |
| `lvc` (V_d) | 27.60 L (RSE 5.8%) | Table 2 |
| `etalcl` | 69.3 %CV -> omega^2 = log(0.693^2 + 1) = 0.392211 | Table 2 and its CV footnote |
| `etalvc` | 22.1 %CV -> omega^2 = log(0.221^2 + 1) = 0.047686 | Table 2 and its CV footnote |
| `addSd` | 0.156 mg/L | Table 2; Results, "Population PK" |
| `propSd` | 0.080 (8.0%) | Table 2; Results, "Population PK" |
| Simulation `QEFF` | `0.77 * CRRT dose * WT / 1000` L/h | Methods, "Simulation-Based Treatment Optimization" |
| Peak time for Cmax | 30 min after the end of the 30-min infusion | Methods, "Simulation-Based Treatment Optimization" |

# The CRRT clearance covariate

`QEFF` is not estimated: the authors calculated a CRRT clearance for each
patient from the circuit settings (Equations 2-8) and added it to the
estimated endogenous clearance. Every value in the Table 1 "CRRT Clearance"
column is reproduced to its printed precision by the post-dilution form
`CRRT dose [mL/kg/h] * weight [kg] * coefficient / 1000`, using the per-patient
sieving coefficient for the Petejova (CVVH) patients and the imputed
saturation coefficient 0.77 for the D'Arcy (CVVHDF) patients. The table below
transcribes Table 1 and checks it; the same formula, with the coefficient
fixed at 0.77, is the one the paper's own dosing simulations use.

```{r table-1}
table1 <- tibble::tribble(
  ~patient, ~study,     ~modality, ~wt, ~dose_mgkg, ~crrt_dose, ~coef, ~clcrrt_pub,
   1L,      "Petejova", "CVVH",     80,  3.0,       45.0,       0.75,  2.7,
   2L,      "Petejova", "CVVH",     94,  2.5,       45.0,       0.82,  3.5,
   3L,      "Petejova", "CVVH",     72,  3.3,       45.0,       0.87,  2.8,
   4L,      "Petejova", "CVVH",     90,  2.7,       45.0,       0.76,  3.1,
   5L,      "Petejova", "CVVH",     90,  2.7,       45.0,       0.78,  3.2,
   6L,      "Petejova", "CVVH",     90,  2.7,       45.0,       0.76,  3.1,
   7L,      "Petejova", "CVVH",    102,  2.4,       45.0,       0.77,  3.5,
   8L,      "D'Arcy",   "CVVHDF",   48,  5.0,       67.7,       0.77,  2.5,
   9L,      "D'Arcy",   "CVVHDF",   68,  5.0,       41.3,       0.77,  2.2,
  10L,      "D'Arcy",   "CVVHDF",   64,  5.0,       47.0,       0.77,  2.3,
  11L,      "D'Arcy",   "CVVHDF",   72,  5.0,       38.6,       0.77,  2.1,
  12L,      "D'Arcy",   "CVVHDF",   56,  5.0,       42.1,       0.77,  1.8,
  13L,      "D'Arcy",   "CVVHDF",   56,  5.0,       36.6,       0.77,  1.6,
  14L,      "D'Arcy",   "CVVHDF",   64,  5.0,       47.2,       0.77,  2.3
) |>
  dplyr::mutate(QEFF = crrt_dose * wt * coef / 1000)

table1 |>
  dplyr::mutate(QEFF = round(QEFF, 3)) |>
  dplyr::select(patient, study, modality, wt, crrt_dose, coef, QEFF, clcrrt_pub) |>
  dplyr::rename(
    "Patient"                  = patient,
    "Study"                    = study,
    "CRRT"                     = modality,
    "Weight (kg)"              = wt,
    "CRRT dose (mL/kg/h)"      = crrt_dose,
    "Sc / Sd"                  = coef,
    "QEFF, recomputed (L/h)"   = QEFF,
    "Table 1 CL_CRRT (L/h)"    = clcrrt_pub
  ) |>
  knitr::kable(caption = "Table 1 of He 2022 and the recomputed CRRT clearance.")

# Table 1 prints one decimal place, so a correct formula lands within 0.05.
stopifnot(max(abs(table1$QEFF - table1$clcrrt_pub)) <= 0.05 + 1e-9)
```

The Discussion quotes a typical total clearance of 63.3 mL/min, which is the
estimated endogenous clearance plus the median CRRT clearance:

```{r typical-cl}
ini_df  <- readModelDb("He_2022_gentamicin")()$iniDf
theta   <- stats::setNames(ini_df$est, ini_df$name)
cl_body <- exp(theta[["lcl"]])
vd      <- exp(theta[["lvc"]])
cl_tot  <- cl_body + stats::median(table1$clcrrt_pub)
c(cl_body_mL_min = cl_body * 1000 / 60, cl_total_mL_min = cl_tot * 1000 / 60)
stopifnot(round(cl_body * 1000 / 60) == 20,        # "only 20 mL/min"
          round(cl_tot * 1000 / 60, 1) == 63.3)    # "typical total clearance of 63.3 mL/min"
```

# Virtual cohort

Observed data are not publicly available. The first simulation replays the
14 Table 1 patients (their weight, mg/kg dose and CRRT clearance), 14 replicates
each, over five q24h doses given as 30-minute infusions, so that each study arm
holds 98 virtual patients.

```{r cohort}
# rxode2's simulation RNG is partitioned per solver thread, so the drawn
# cohort is not reproducible across machines with different thread counts.
# Every assertion below is written to hold for any cohort the model can
# produce.
rxode2::rxSetSeed(2022)

N_REP   <- 14L
INF_DUR <- 0.5   # h, Table 1 / Methods
TAU     <- 24    # h
N_DOSE  <- 5L

obs_times <- sort(unique(round(c(
  seq(0, N_DOSE * TAU, by = 0.5),
  (0:(N_DOSE - 1)) * TAU + 1      # the 30-min post-infusion peak
), 6)))

events <- table1 |>
  tidyr::crossing(rep = seq_len(N_REP)) |>
  dplyr::mutate(id = dplyr::row_number()) |>
  dplyr::group_by(id) |>
  dplyr::group_modify(function(d, key) {
    ev <- rxode2::et(amt = d$dose_mgkg * d$wt, dur = INF_DUR, ii = TAU,
                     addl = N_DOSE - 1L, cmt = "central") |>
      rxode2::et(obs_times, cmt = "central")
    as.data.frame(ev) |>
      dplyr::select(-dplyr::any_of("id")) |>
      dplyr::mutate(QEFF = d$QEFF, study = d$study, WT = d$wt,
                    dose_mg = d$dose_mgkg * d$wt)
  }) |>
  dplyr::ungroup() |>
  as.data.frame()

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])),
          max(table(unique(events[, c("id", "study")])$study)) <= 200)
dplyr::count(dplyr::distinct(events, id, study), study)
```

# Simulation

```{r simulate}
mod <- readModelDb("He_2022_gentamicin")

sim <- rxode2::rxSolve(
  mod, events = events,
  keep = c("study", "QEFF", "dose_mg"),
  # Tight tolerances: the closed-form gate below compares the ODE solve with
  # the analytic infusion solution on the same drawn parameters.
  rtol = 1e-10, atol = 1e-12
) |>
  as.data.frame() |>
  dplyr::mutate(time = round(time, 6))
```

## The ODE solve reproduces the closed-form solution

For a one-compartment model with a 30-minute infusion repeated every 24 h,
the concentration is a superposition of single-infusion curves. Both sides of
this comparison use the same drawn clearance and volume, so the only
difference is solver error. It also shows that `QEFF` reaches the solve: the
CRRT arm is roughly two-thirds of the total clearance, so dropping it would
change concentrations by far more than the bound.

```{r closed-form}
# Written with pmin()/pmax() rather than ifelse(): ifelse() takes its length
# from the test, so a scalar time with vectors of clearances would silently
# return a single value.
c_single <- function(s, dose, cl, v) {
  k <- cl / v
  rate <- dose / INF_DUR
  t_in  <- pmin(pmax(s, 0), INF_DUR)   # time spent infusing so far
  t_out <- pmax(s - INF_DUR, 0)        # time since the end of the infusion
  rate / cl * (1 - exp(-k * t_in)) * exp(-k * t_out)
}
c_multi <- function(t, dose, cl, v, n = N_DOSE, tau = TAU) {
  out <- 0
  for (j in 0:(n - 1)) out <- out + c_single(t - j * tau, dose, cl, v)
  out
}

cf <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(closed = c_multi(time, dose_mg, cl, vc)) |>
  dplyr::filter(closed > 0)

stopifnot(nrow(cf) > 0, all(abs(cf$cl - (cf$cl_body + cf$QEFF)) < 1e-10))
max(abs(cf$Cc / cf$closed - 1))
stopifnot(max(abs(cf$Cc / cf$closed - 1)) < 1e-6)
```

# Replicate published figures

## Simulated counterpart of Figure 1

Figure 1 of He 2022 is a scatter plot of the observed concentrations after
the dose for the 14 patients. The plot below shows the simulated median and
5th-95th percentiles of the first dosing interval for each study, with the
same regimens and CRRT clearances. The D'Arcy patients received about twice
the mg/kg dose (5 versus about 2.7 mg/kg) but were lighter, and the volume of
distribution is not weight-scaled, so their absolute doses, and hence their
peaks, are only about a third higher (checked on the NCA below).

```{r figure-1}
sim |>
  dplyr::filter(time <= TAU, !is.na(Cc)) |>
  dplyr::group_by(study, time) |>
  dplyr::summarise(
    q05 = stats::quantile(Cc, 0.05), q50 = stats::median(Cc),
    q95 = stats::quantile(Cc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(time, q50, colour = study, fill = study)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.2, colour = NA) +
  geom_line() +
  labs(x = "Time after dose (h)", y = "Gentamicin (mg/L)", colour = NULL, fill = NULL,
       title = "First dosing interval, Table 1 patients replayed",
       caption = "Simulated counterpart of Figure 1 of He 2022 (median and 90% interval).")
```

# PKNCA validation

PKNCA on the first dosing interval of the stochastic simulation, grouped by
study, followed by a typical-value check: for a single dose,
`AUC0-inf * CL = Dose` exactly, and the terminal half-life is
`log(2) * V / CL`.

```{r pknca}
conc_first <- sim |>
  dplyr::filter(!is.na(Cc), time <= TAU) |>
  dplyr::select(id, time, Cc, study)
dose_first <- events |>
  dplyr::filter(evid != 0, time == 0) |>
  dplyr::select(id, time, amt, study)

o_conc <- PKNCA::PKNCAconc(conc_first, Cc ~ time | study + id)
o_dose <- PKNCA::PKNCAdose(dose_first, amt ~ time | study + id)
o_data <- PKNCA::PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(start = 0, end = TAU, cmax = TRUE, tmax = TRUE,
                         auclast = TRUE, cmin = TRUE)
)
nca_first <- PKNCA::pk.nca(o_data)
summary(nca_first)

# Peak ratio between the studies follows the ratio of mean absolute doses.
cmax_by_study <- as.data.frame(nca_first$result) |>
  dplyr::filter(PPTESTCD == "cmax") |>
  dplyr::group_by(study) |>
  dplyr::summarise(cmax = stats::median(PPORRES))
dose_ratio <- with(table1, mean((dose_mgkg * wt)[study == "D'Arcy"]) /
                             mean((dose_mgkg * wt)[study == "Petejova"]))
cmax_ratio <- cmax_by_study$cmax[cmax_by_study$study == "D'Arcy"] /
  cmax_by_study$cmax[cmax_by_study$study == "Petejova"]
c(dose_ratio = dose_ratio, cmax_ratio = cmax_ratio)
# The medians of 98 subjects per study carry a few percent of sampling noise
# (22% CV on the volume), and the higher CRRT clearance of the Petejova
# patients lowers their peak slightly more; 25% still fails on a twofold
# dose or unit error.
stopifnot(abs(cmax_ratio / dose_ratio - 1) < 0.25)
```

```{r pknca-typical}
# Typical patient: 72 kg (Table 1 median), 5 mg/kg single dose, CRRT
# clearance at the Table 1 median of 2.6 L/h.
ev_tv <- rxode2::et(amt = 5 * 72, dur = INF_DUR, cmt = "central") |>
  rxode2::et(round(c(0, 0.25, 0.5, 1, seq(2, 96, by = 2)), 6), cmt = "central") |>
  as.data.frame() |>
  dplyr::mutate(QEFF = 2.6, id = 1L, treatment = "5 mg/kg, typical")

sim_tv <- rxode2::rxSolve(rxode2::zeroRe(mod), events = ev_tv,
                          keep = "treatment", rtol = 1e-10, atol = 1e-12) |>
  as.data.frame()

conc_tv <- sim_tv |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(id = 1L) |>
  dplyr::select(id, time, Cc, treatment)
dose_tv <- ev_tv |>
  dplyr::filter(evid != 0) |>
  dplyr::select(id, time, amt, treatment)

nca_tv <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_tv, Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(dose_tv, amt ~ time | treatment + id),
  intervals = data.frame(start = 0, end = Inf, cmax = TRUE,
                         aucinf.obs = TRUE, half.life = TRUE)
))
res_tv <- as.data.frame(nca_tv$result)
res_tv[, c("PPTESTCD", "PPORRES")]

auc_inf <- res_tv$PPORRES[res_tv$PPTESTCD == "aucinf.obs"]
thalf   <- res_tv$PPORRES[res_tv$PPTESTCD == "half.life"]
cl_tv   <- cl_body + 2.6
c(auc_x_cl = auc_inf * cl_tv, dose = 5 * 72,
  half_life = thalf, closed_half_life = log(2) * vd / cl_tv)
# Trapezoidal NCA on a 2-h grid: 1% agreement is well above the discretisation
# error and well below what a mis-transcribed CL, V or QEFF would produce.
stopifnot(abs(auc_inf * cl_tv / (5 * 72) - 1) < 0.01,
          abs(thalf / (log(2) * vd / cl_tv) - 1) < 0.01)
```

He 2022 reports no NCA table, so there is no published NCA to compare
against. The half-life of about 5 h and the volume of 0.38 L/kg at the 72 kg
median weight follow directly from Table 2 (the Discussion quotes 0.39 L/kg).

# Tables 3-5: target attainment and toxicity at steady state

The paper simulated 15 scenarios (3-7 mg/kg q24h as a 30-minute infusion,
crossed with CRRT doses of 30, 40 and 50 mL/kg/h) in 10,000 MIMIC-III patients
with weight restricted to 48-102 kg, each patient's CRRT clearance being
`0.77 * CRRT dose * weight / 1000`. It reports, after the fifth dose, the
percentage of patients with AUC0-24h/MIC > 100, Cmax/MIC > 10 (Cmax 30 min
after the end of infusion) and Cmin below 1 or 2 mg/L.

## Cells that the model structure forces to zero

At steady state `AUC0-24h = Dose / CL`. With the dose in mg/kg and the CRRT
clearance proportional to weight, `AUC0-24h / MIC > 100` needs

`CL_body * exp(eta) < WT * (dose_mg_per_kg / (100 * MIC) - 0.77 * CRRT_dose / 1000)`.

Whenever the bracket is zero or negative, no patient of any weight, and no
draw of the random effect, can reach the target. That set of cells is fixed
by the model structure and the simulation design alone, and must be exactly
the set of cells the paper prints as `0`.

```{r structural-zeros}
published <- tibble::tribble(
  ~crrt, ~dose, ~auc_mic1, ~auc_mic2, ~cmax_mic1, ~cmax_mic2, ~cmin_lt1, ~cmin_lt2,
  30, 3,  9.85, 0,    26.2, 0.03, 71.1, 98.3,
  30, 4, 53.1,  0,    72.4, 0.88, 55.3, 90.3,
  30, 5, 78.5,  0.08, 93.5, 7.70, 44.3, 80.8,
  30, 6, 89.5,  9.85, 98.9, 26.2, 37.2, 71.1,
  30, 7, 94.9, 32.1,  99.8, 51.7, 31.4, 62.3,
  40, 3,  0,    0,    21.6, 0.02, 92.6, 99.9,
  40, 4, 19.9,  0,    65.6, 0.7,  81.9, 99.5,
  40, 5, 60.7,  0,    90.4, 6.07, 71.6, 97.2,
  40, 6, 81.9,  0,    98.2, 21.6, 61.9, 92.6,
  40, 7, 91.1,  1.96, 99.7, 44.3, 54.3, 87.5,
  50, 3,  0,    0,    19.2, 0.04, 99.3, 100,
  50, 4,  0.02, 0,    61.0, 0.61, 95.8, 99.9,
  50, 5, 30.4,  0,    87.9, 5.37, 90.3, 99.8,
  50, 6, 67.4,  0,    97.4, 19.2, 84.0, 99.3,
  50, 7, 84.8,  0,    99.6, 40.1, 78.0, 98.1
)

auc_cells <- published |>
  dplyr::select(crrt, dose, auc_mic1, auc_mic2) |>
  tidyr::pivot_longer(c(auc_mic1, auc_mic2), names_to = "cell", values_to = "pub") |>
  dplyr::mutate(
    mic        = ifelse(cell == "auc_mic1", 1, 2),
    bracket    = dose / (100 * mic) - 0.77 * crrt / 1000,
    impossible = bracket <= 0
  )

auc_cells |>
  dplyr::filter(impossible | pub == 0) |>
  dplyr::select(crrt, dose, mic, bracket, pub)

stopifnot(identical(auc_cells$impossible, auc_cells$pub == 0))

# The three smallest non-zero published cells are the three cells with the
# smallest positive bracket.
possible <- dplyr::filter(auc_cells, !impossible)
stopifnot(identical(order(possible$bracket)[1:3], order(possible$pub)[1:3]))
```

Every printed zero in the two AUC columns (`r sum(auc_cells$pub == 0)` of the
`r nrow(auc_cells)` cells) is a structurally impossible cell, and every
structurally impossible cell is printed as zero. This pins the simulated CRRT clearance to
`0.77 * CRRT dose * weight / 1000`, without the pre-dilution factor. The
smallest positive cells (0.02%, 0.08%, 1.96%) sit exactly where the bracket
is just above zero.

## All 90 cells

The non-zero cells depend on the weight distribution of the MIMIC-III
subset, which the paper does not report. The maintainers assumed a normal
distribution with mean 80 kg and SD 20 kg, truncated to 48-102 kg by
rejection. Because the model is one-compartment with a single IV infusion
route, each patient's concentrations have the closed form above (confirmed
against the ODE solve), so the cells are evaluated on 20,000 draws per
scenario with R's own generator. Unlike the rxode2 generator, this one
reproduces across machines and thread counts. The same random numbers are
reused across scenarios, as in any common-random-numbers design.

```{r tables-3-5}
set.seed(20220105)
N_MC <- 20000L
draw_wt <- function(n) {
  w <- numeric(0)
  while (length(w) < n) {
    x <- stats::rnorm(n, 80, 20)
    w <- c(w, x[x >= 48 & x <= 102])
  }
  w[seq_len(n)]
}
wt_mc  <- draw_wt(N_MC)
eta_cl <- stats::rnorm(N_MC, 0, sqrt(theta[["etalcl"]]))
eta_v  <- stats::rnorm(N_MC, 0, sqrt(theta[["etalvc"]]))

# Fifth dose: t in [96, 120] h. AUC over that interval from the mass balance
# (amount in - amount out) / CL.
scenario <- function(crrt, dose, t_peak = 1) {
  cl   <- cl_body * exp(eta_cl) + 0.77 * crrt * wt_mc / 1000
  v    <- vd * exp(eta_v)
  amt  <- dose * wt_mc
  t0   <- (N_DOSE - 1) * TAU
  cmax <- c_multi(t0 + t_peak, amt, cl, v)
  cmin <- c_multi(t0 + TAU, amt, cl, v)
  a0   <- v * c_multi(t0, amt, cl, v)
  auc  <- (amt - v * cmin + a0) / cl
  tibble::tibble(
    auc_mic1  = 100 * mean(auc > 100),  auc_mic2  = 100 * mean(auc / 2 > 100),
    cmax_mic1 = 100 * mean(cmax > 10),  cmax_mic2 = 100 * mean(cmax / 2 > 10),
    cmin_lt1  = 100 * mean(cmin < 1),   cmin_lt2  = 100 * mean(cmin < 2)
  )
}

simulated <- published |>
  dplyr::select(crrt, dose) |>
  dplyr::rowwise() |>
  dplyr::mutate(scenario(crrt, dose)) |>
  dplyr::ungroup()

cmp <- dplyr::inner_join(
  tidyr::pivot_longer(simulated, -c(crrt, dose), values_to = "model"),
  tidyr::pivot_longer(published, -c(crrt, dose), values_to = "paper"),
  by = c("crrt", "dose", "name")
) |>
  dplyr::mutate(diff = model - paper)

cmp |>
  dplyr::mutate(dplyr::across(c(model, paper, diff), \(x) round(x, 1)),
                name = factor(name, levels = names(published)[-(1:2)])) |>
  dplyr::arrange(name, crrt, dose) |>
  dplyr::rename(
    "CRRT dose (mL/kg/h)"   = crrt,
    "Gentamicin (mg/kg q24h)" = dose,
    "Index"                 = name,
    "This model (%)"        = model,
    "He 2022 (%)"           = paper,
    "Difference (pp)"       = diff
  ) |>
  knitr::kable(caption = "Replicates Tables 3, 4 and 5 of He 2022 (steady state, after the fifth dose).")
```

```{r tables-3-5-gates}
by_index <- cmp |>
  dplyr::mutate(kind = sub("_.*", "", name)) |>
  dplyr::group_by(kind) |>
  dplyr::summarise(median_abs = stats::median(abs(diff)),
                   max_abs = max(abs(diff)), median_signed = stats::median(diff))
by_index

auc_cmin <- cmp |> dplyr::filter(!grepl("^cmax", name))
cmax     <- cmp |> dplyr::filter(grepl("^cmax", name))
stopifnot(
  # AUC and trough cells are within a few points of the paper despite the
  # assumed weight distribution (observed maximum about 2 points).
  max(abs(auc_cmin$diff)) < 5,
  # Peak cells: see the deviations section. The shortfall is systematic but
  # bounded; a mis-transcribed volume would move these by tens of points.
  stats::median(abs(cmax$diff)) < 6,
  max(abs(cmax$diff)) < 15
)

# The same peak cells read at the END of the infusion (t = 0.5 h) instead.
cmax_eoi <- published |>
  dplyr::select(crrt, dose) |>
  dplyr::rowwise() |>
  dplyr::mutate(scenario(crrt, dose, t_peak = INF_DUR)) |>
  dplyr::ungroup() |>
  dplyr::select(crrt, dose, cmax_mic1, cmax_mic2) |>
  tidyr::pivot_longer(-c(crrt, dose), values_to = "model_eoi") |>
  dplyr::inner_join(cmax, by = c("crrt", "dose", "name")) |>
  dplyr::mutate(diff_eoi = model_eoi - paper)
c(median_signed_30min = stats::median(cmax$diff),
  median_signed_end_of_infusion = stats::median(cmax_eoi$diff_eoi))
# The paper's peak cells lie between the two sampling times.
stopifnot(stats::median(cmax$diff) < 0, stats::median(cmax_eoi$diff_eoi) > 0)
```

The AUC and trough cells agree with the paper to within about two percentage
points. The Cmax cells run systematically low by up to about nine points
(median about three), with the largest gaps in the steep middle of the
attainment curve. The model's peak is read exactly 30 minutes after the end
of the infusion, as the Methods define it. Reading it at the end of the
infusion instead moves the simulated cells to a few points *above* the paper,
so the paper's peak sits between the two. Heavier patients receive a larger
absolute dose into a volume that is not weight-scaled, so the peak cells are
also the ones most sensitive to the unreported weight distribution.

# Assumptions and deviations

* **CRRT clearance as data.** `QEFF` is the per-patient CRRT clearance in L/h,
  supplied as a covariate and not estimated, exactly as in the source model.
  Users must compute it from the CRRT prescription (Equations 2-8 of He 2022;
  for the paper's simulations `0.77 * CRRT dose [mL/kg/h] * weight [kg] / 1000`).
* **Four patients with CL_body set to zero.** For four patients (1, 6, 13 and
  14 in Table 1) the non-compartmental ratio `CL_body / CL_total` was below 0.05
  and the authors "assumed [CL_total] to be equal to CL_CRRT" during model
  development. The paper does not say how this was coded in NONMEM. It is a
  property of the fitting data, not of the final model, and is not
  represented in the model file. Every simulated patient carries the
  estimated endogenous clearance with its random effect, as in the paper's
  own simulations.
* **Weight distribution for Tables 3-5.** The MIMIC-III weight distribution
  after truncation to 48-102 kg is not reported. The maintainers assumed a
  normal distribution (mean 80 kg, SD 20 kg) truncated by rejection. The
  AUC-zero identity above does not depend on this choice.
* **Steady state.** "Steady state" is taken as the fifth dosing interval
  (96-120 h), following the Methods ("after the fifth dosing").
* **Residual error.** Additive 0.156 mg/L plus proportional 8.0%, encoded as
  nlmixr2's `add() + prop()`. The paper does not state whether the two
  components combine on the SD or the variance scale. The PTA tables use
  predicted concentrations without residual error; the exact equalities in
  the published tables (for example, 3 mg/kg at MIC 1 and 6 mg/kg at MIC 2
  both give 9.85% for AUC and 26.2% for Cmax at 30 mL/kg/h) hold only
  without an additive error.
* **Supplement.** The first-day tables (Tables S1-S3), the pcVPC (Figure S1)
  and the accumulation profiles (Figure S2) are in the online supplement and
  were not used; they carry no model parameters.
* **Screened covariates.** Body weight, age, sex and CRRT modality were
  tested and not retained. They are recorded in `covariatesDataExcluded`.
