---
title: "Mycophenolic acid (Heida 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mycophenolic acid (Heida 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Heida_2024_mycophenolic_acid"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1007/s00228-024-03743-0>
* Supplementary material (NONMEM control stream of the final model):
  <https://doi.org/10.1007/s00228-024-03743-0> (Electronic Supplementary Material 1)

Heida and colleagues developed a population PK model for mycophenolic acid
(MPA) from routine therapeutic-drug-monitoring data in pediatric kidney
transplant recipients at Radboudumc, externally evaluated it against a second
Dutch cohort, and used it to build a weight-banded starting-dose scheme for
mycophenolate mofetil (MMF).

**Unit convention.** This is the single most important thing to know before
using the model. Doses are entered in milligrams of *mycophenolate mofetil*
(the prodrug), while concentrations are milligrams per litre of *mycophenolic
acid* (Methods, Model development: "the dose was entered in MMF equivalents
... and the dependent variable ... was entered as MPA equivalents"). `CL/F`
and the volumes are therefore apparent values referenced to the MMF dose, and
already absorb the 0.739 MMF-to-MPA molecular-weight ratio. The Discussion
makes this explicit when comparing against the literature: the authors
multiply their estimates by 0.739 before placing them alongside MPA-dosed
models.

# Population

The model-development dataset is 30 pediatric kidney transplant recipients
treated with oral MMF (CellCept) between June 2016 and April 2023 at the
Amalia Children's Hospital, Radboudumc, Nijmegen. Median age was 13 years
(range 4 to 18), median weight 38.5 kg (12.9 to 79.9), median height 149 cm
(95 to 193), median body surface area 1.3 m2 (0.58 to 2.1) and median serum
albumin 34 g/L (24 to 42); 12 of 30 were female (Table 1). Immunosuppressive
co-medication was tacrolimus in 25 patients and everolimus in 5, with
prednisone in 13.3%; neither partner is known to affect MMF PK. Patients on
ciclosporin were excluded, because ciclosporin inhibits MPA enterohepatic
recirculation, so the model should not be extrapolated to ciclosporin
co-treatment.

Sampling was mostly early after transplantation (median post-transplant time
9.5 days, range 2 to 3058) and yielded 266 MPA plasma concentrations: 20 full
PK curves, 24 limited-sampling curves and 25 trough levels, a median of 9.5
observations per patient. MPA was assayed by a validated EMIT immunoassay
(Roche Cobas c502), which over-reads MPA relative to HPLC through
cross-reactivity with the acyl-glucuronide metabolite, so the model is
calibrated to EMIT-scale concentrations.

The same information is available programmatically via
`readModelDb("Heida_2024_mycophenolic_acid")()$population`.

# Source trace

Per-parameter origins are recorded as in-file comments beside each `ini()`
entry in `inst/modeldb/specificDrugs/Heida_2024_mycophenolic_acid.R`. They are
collected here for review. "Control stream" refers to Electronic
Supplementary Material 1, which contains the final model's NONMEM code and,
per the Results, its final parameter estimates.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL/F) | 16.0 L/h | Table 2 theta1 (95% CI 10.3-20.4); Eq. 1; control stream `$THETA 1` |
| `lvc` (Vc/F) | 24.9 L | Table 2 (95% CI 6.53-45.8); Eq. 2; control stream `$THETA 2` |
| `lvp` (Vp/F) | 1590 L | Table 2 (95% CI 651-2994); Eq. 3; control stream `$THETA 3` |
| `lq` (Q/F) | 36.2 L/h | Table 2 (95% CI 25.8-49.6); Eq. 4; control stream `$THETA 4` |
| `lktr` (ktr) | 1.48 1/h | Table 2 (95% CI 1.15-1.84); Eq. 5; control stream `$THETA 5` |
| `lfdepot` (F) | 1 (fixed) | Control stream `$PK` `F1=1*EXP(IOV)`; no absolute-bioavailability data (Methods) |
| `e_wt_cl`, `e_wt_q` | 0.75 (fixed) | Methods, Model development; Eqs. 1 and 4; control stream `ALLOCL=(WT/70)**0.75` |
| `e_wt_vc`, `e_wt_vp` | 1 (fixed) | Methods, Model development; Eqs. 2 and 3; control stream `ALLOV=(WT/70)` |
| `e_wt_ktr` | -0.25 (fixed) | Methods, Model development; Eq. 5; control stream `ALLOK=(WT/70)**(-0.25)` |
| `e_alb_cl` | -2.49 | Table 2 theta2 (95% CI -3.82 to -1.41); control stream `COVALB=(ALB/34)**THETA(6)`, `$THETA 6` |
| `etalcl` | 0.13889 | Table 2 IIV Cl/F 38.6% CV; Methods CV formula; control stream `$OMEGA 1` = 0.139 |
| `etalvc` | 2.41955 | Table 2 IIV Vc/F 320% CV; control stream `$OMEGA 2` = 2.42 |
| `etalq` | 0.33997 | Table 2 IIV Q/F 63.6% CV; control stream `$OMEGA 4` = 0.337 |
| `etaiov_fdepot_1..10` | 0.19267 | Table 2 IOV F 46.1% CV; control stream `$OMEGA BLOCK(1) 0.19` + nine `SAME` |
| `propSd` | 0.473 | Table 2 residual error 47.3%; control stream `$SIGMA` = 0.223 |
| Erlang absorption, `depot -> transit1 -> central` | n/a | Results, Model development; Fig. 2; control stream `$MODEL` `COMP=(DOSE)/(TRAN)/(CENTRAL)/(PERIPHERAL)` with `K12=K23=KTR` |
| Two-compartment disposition | n/a | Results, Model development; control stream `K34=Q/V3`, `K43=Q/V4`, `K30=CL/V3` |
| Proportional residual error | n/a | Results, Model development; control stream `$ERROR` `Y=IPRED+IPRED*ERR(1)` |

```{r source-trace-cv}
# The paper states its variance transform explicitly (Methods, Model
# development): CV(%) = sqrt(exp(omega^2) - 1) * 100. Confirm that the
# variances shipped in ini() reproduce the Table 2 percentages exactly.
omega_to_cv <- function(v) sqrt(exp(v) - 1) * 100

cv_check <- tibble::tribble(
  ~Term,                              ~Variance,  ~`Table 2 CV%`,
  "IIV CL/F",                         0.13889,    38.6,
  "IIV Vc/F",                         2.41955,    320,
  "IIV Q/F",                          0.33997,    63.6,
  "IOV F",                            0.19267,    46.1
) |>
  dplyr::mutate(`Back-transformed CV%` = omega_to_cv(Variance))

knitr::kable(cv_check, digits = c(NA, 5, 1, 1),
             caption = "ini() variances back-transform to the Table 2 CVs.")

# Deterministic arithmetic, so this is asserted tightly.
stopifnot(all(abs(cv_check$`Back-transformed CV%` - cv_check$`Table 2 CV%`) < 0.06))
```

# Structural checks

Before any cohort simulation, three deterministic properties of the
implemented model are checked against arithmetic that does not use the model.

```{r structural}
mod <- readModelDb("Heida_2024_mycophenolic_acid")

# Typical-value parameters at the cohort-median weight and the reference
# albumin, computed independently of the model from Eqs. 1-5.
eq_params <- function(WT, ALB = 34) {
  list(
    cl  = 16.0 * (WT / 70)^0.75 * (ALB / 34)^-2.49,
    vc  = 24.9 * (WT / 70),
    vp  = 1590 * (WT / 70),
    q   = 36.2 * (WT / 70)^0.75,
    ktr = 1.48 * (WT / 70)^-0.25
  )
}

p <- eq_params(38.5)
k10 <- p$cl / p$vc; k12 <- p$q / p$vc; k21 <- p$q / p$vp
ssum <- k10 + k12 + k21; sprod <- k10 * k21
beta  <- (ssum - sqrt(ssum^2 - 4 * sprod)) / 2
alpha <- (ssum + sqrt(ssum^2 - 4 * sprod)) / 2

structural <- tibble::tibble(
  Quantity = c("Mean absorption time 2/ktr (h)",
               "Distribution half-life log(2)/alpha (h)",
               "Terminal half-life log(2)/beta (h)",
               "Time to 95% of steady state, 3*log(2)/beta (h)"),
  Value = c(2 / p$ktr, log(2) / alpha, log(2) / beta, 3 * log(2) / beta)
)
knitr::kable(structural, digits = 2,
             caption = "Derived disposition timescales at the cohort-median weight of 38.5 kg.")
```

Two consequences are worth stating plainly, because they are properties of
the published model rather than of this implementation.

* Absorption is Erlang with **shape 2**, not shape 1. The control stream's
  `$MODEL` block declares `COMP=(DOSE)` and `COMP=(TRAN)` ahead of
  `COMP=(CENTRAL)` and sets `K12 = K23 = KTR`, so the dose crosses **two**
  sequential first-order steps of equal rate before reaching plasma. The
  paper describes this as "a single transition compartment", counting the
  transition compartment only and not the dosing compartment. Mean absorption
  time is `2/ktr`, giving a typical Tmax near 1 h, which matches the dense
  band of early observations in the paper's Figure 1.
* The terminal half-life is long -- about 86 h at the cohort-median weight --
  because `Vp/F` is 1590 L with a slow return rate `Q/Vp`. The paper itself
  flags this estimate as an outlier ("the Vp/F estimation of 1590 L was
  higher than previously reported values of 16.8 to 411.0 L"). The practical
  consequence, quantified in the steady-state section below, is that AUC0-12h
  approaches its steady-state value only after several weeks of dosing.

## Body surface area reproduces Table 1

The paper computed body surface area with the Dubois formula (Methods, Study
population). Applying it to the paired weight and height rows of Table 1
should reproduce the Table 1 body-surface-area row. This is an arithmetic
check on the demographic table, and it simultaneously establishes that the
minimum, median and maximum of the weight and height rows belong to the same
patients -- which is what lets the virtual cohort below derive height from
weight.

```{r bsa-check}
dubois <- function(wt, ht) 0.007184 * wt^0.425 * ht^0.725

bsa_check <- tibble::tibble(
  Statistic = c("Minimum", "Median", "Maximum"),
  `Weight (kg)` = c(12.9, 38.5, 79.9),
  `Height (cm)` = c(95, 149, 193),
  `Published BSA (m2)` = c(0.58, 1.3, 2.1)
) |>
  dplyr::mutate(`Dubois BSA (m2)` = dubois(`Weight (kg)`, `Height (cm)`))

knitr::kable(bsa_check, digits = 3,
             caption = "Dubois body surface area from the Table 1 weight and height rows.")

# Deterministic: the published BSA row is printed to 2 significant figures,
# so agreement to within half a unit in the last printed digit is exact
# agreement.
stopifnot(all(abs(bsa_check$`Dubois BSA (m2)` - bsa_check$`Published BSA (m2)`) < 0.03))
```

## Steady-state AUC reproduces dose divided by clearance

The dosing scheme in Table 3 was built from the steady-state identity
`dose = clearance * AUC` (Methods, Development of an improved starting dose).
For a linear model, superposition makes that identity exact in the form
`AUC0-tau at steady state = AUC0-inf after a single dose = dose / CL`, so the
single-dose form is used here: it is the same arithmetic without needing the
simulation to converge on steady state. It checks the whole implemented chain
at once -- a mis-transcribed clearance, allometric exponent, albumin
exponent, reference value, compartment connection or dose unit all break it.

```{r auc-identity-gate}
trapz <- function(x, y) sum(diff(x) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)

# The terminal half-life reaches ~150 h at the low-clearance corner of the
# covariate space, so the grid runs to 4000 h (>25 half-lives) with fine
# early sampling to resolve the absorption and distribution phases.
auc_grid <- unique(c(seq(0, 24, by = 0.05), seq(24, 240, by = 0.5),
                     seq(240, 4000, by = 10)))

single_dose_auc <- function(WT, ALB, dose) {
  ev <- rxode2::et(amt = dose, cmt = "depot") |>
    rxode2::et(auc_grid, cmt = "central")
  ev <- as.data.frame(ev)
  ev$WT <- WT; ev$ALB <- ALB; ev$OCC <- 1
  out <- rxode2::rxSolve(mod, ev, omega = NA, sigma = NA,
                         returnType = "data.frame")
  out <- out[!is.na(out$Cc), ]
  trapz(out$time, out$Cc)
}

auc_check <- tidyr::expand_grid(WT = c(15, 30, 50, 75), ALB = c(24, 34, 42)) |>
  dplyr::mutate(dose = 500) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    `Simulated AUC0-inf` = single_dose_auc(WT, ALB, dose),
    `Dose / CL` = dose / eq_params(WT, ALB)$cl
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(Ratio = `Simulated AUC0-inf` / `Dose / CL`)

knitr::kable(
  auc_check |> dplyr::rename("Weight (kg)" = WT, "Albumin (g/L)" = ALB,
                             "Dose (mg)" = dose),
  digits = c(0, 0, 0, 2, 2, 5),
  caption = "Single-dose AUC0-inf from the ODE system against the closed-form dose / clearance."
)

# Deterministic identity, limited only by the trapezoidal grid and the
# truncation at 4000 h. Realised max |ratio - 1| was 6.2e-05; 1e-3 leaves an
# order of magnitude of headroom yet still goes red on any transcription
# error, which moves the ratio by percent to tens of percent.
stopifnot(max(abs(auc_check$Ratio - 1)) < 1e-3)
```

## How long steady state takes

The identity above assumes steady state. Because the terminal half-life is
about 86 h, the assumption is reached slowly. The following traces AUC0-12h
across successive dosing intervals from the first dose onward. This is not a
published result -- it is a property of the model that a user applying the
Table 3 scheme immediately after transplantation should know.

```{r approach-to-ss}
WT_med <- 38.5; ALB_med <- 34; dose_med <- 500
n_int <- 120                                   # 120 intervals of 12 h = 60 days
ev <- rxode2::et(amt = dose_med, cmt = "depot", ii = 12, addl = n_int - 1) |>
  rxode2::et(seq(0, 12 * n_int, by = 0.25), cmt = "central")
ev <- as.data.frame(ev)
ev$WT <- WT_med; ev$ALB <- ALB_med; ev$OCC <- 1
approach <- rxode2::rxSolve(mod, ev, omega = NA, sigma = NA,
                            returnType = "data.frame")
approach <- approach[!is.na(approach$Cc), ]

auc_ref <- dose_med / eq_params(WT_med, ALB_med)$cl
interval_auc <- vapply(seq_len(n_int), function(i) {
  w <- approach[approach$time >= 12 * (i - 1) & approach$time <= 12 * i, ]
  trapz(w$time, w$Cc)
}, numeric(1))

tibble::tibble(day = (seq_len(n_int) * 12) / 24,
               pct = 100 * interval_auc / auc_ref) |>
  ggplot(aes(day, pct)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 95, linetype = "dashed") +
  labs(x = "Day of twice-daily dosing", y = "AUC0-12h as % of steady state",
       title = "AUC0-12h approaches steady state over several weeks",
       caption = "Typical 38.5 kg patient, albumin 34 g/L, 500 mg q12h. Dashed line: 95% of steady state.")

day95 <- min(which(interval_auc / auc_ref >= 0.95)) * 12 / 24
cat(sprintf("AUC0-12h first reaches 95%% of its steady-state value on day %.1f\n", day95))

# Deterministic (no random effects), and the claim being locked in is that
# accumulation is slow rather than a precise day count.
stopifnot(day95 > 7, day95 < 45, interval_auc[1] / auc_ref < 0.8)
```

# Virtual cohort

Original patient-level data are not publicly available. The cohort below
approximates Table 1 of the paper: weight lognormal about the median 38.5 kg
and truncated to the observed 12.9 to 79.9 kg range, height derived from
weight, and serum albumin drawn from the distribution the paper used for its
own dosing simulations (mean 33 g/L, CV 21%; Methods, Development of an
improved starting dose) and truncated to the observed 24 to 42 g/L range.

```{r cohort}
# set.seed() seeds R's RNG for the covariate draws below. It does NOT seed
# rxode2's simulation RNG, and rxode2's streams are partitioned per solver
# thread, so the etas differ between a 16-thread workstation and a 2-core CI
# runner. Every assertion downstream is written to hold for any cohort the
# model can produce; see pattern 12 of known-vignette-failure-patterns.md.
# The RNG kind is pinned explicitly so the covariate draws -- which the gated
# typical-value comparison depends on -- stay identical across R versions.
set.seed(20240817, kind = "Mersenne-Twister", normal.kind = "Inversion")

n_sub <- 200

rtrunc_lnorm <- function(n, med, sdlog, lo, hi) {
  x <- rlnorm(n, meanlog = log(med), sdlog = sdlog)
  pmin(pmax(x, lo), hi)
}
rtrunc_norm <- function(n, mean, sd, lo, hi) {
  pmin(pmax(rnorm(n, mean, sd), lo), hi)
}

# Height from weight. The exponent and coefficient are the log-log fit through
# the three PAIRED (weight, height) anchors of Table 1, whose pairing is
# established by the body-surface-area check above.
ht_fit <- lm(log(c(95, 149, 193)) ~ log(c(12.9, 38.5, 79.9)))
ht_from_wt <- function(wt) exp(coef(ht_fit)[1] + coef(ht_fit)[2] * log(wt))

subjects <- tibble::tibble(
  id  = seq_len(n_sub),
  WT  = rtrunc_lnorm(n_sub, med = 38.5, sdlog = 0.43, lo = 12.9, hi = 79.9),
  ALB = rtrunc_norm(n_sub, mean = 33, sd = 33 * 0.21, lo = 24, hi = 42)
) |>
  dplyr::mutate(
    HT  = ht_from_wt(WT),
    BSA = dubois(WT, HT),
    # Licensed regimen: 1200 mg/m2 per day split over two doses, rounded to
    # the 250 mg tablet (Methods, Development of an improved starting dose).
    dose_licensed = pmax(250, round(600 * BSA / 250) * 250),
    # Table 3 weight-banded model-informed starting dose, administered twice
    # daily.
    dose_optimized = dplyr::case_when(
      WT < 20 ~ 250,
      WT < 40 ~ 500,
      WT < 60 ~ 750,
      TRUE    ~ 1000
    ),
    weight_band = dplyr::case_when(
      WT < 20 ~ "<20 kg", WT < 40 ~ "20-39 kg",
      WT < 60 ~ "40-59 kg", TRUE ~ "60-90 kg"
    )
  )

knitr::kable(
  subjects |>
    dplyr::summarise(
      dplyr::across(c(WT, HT, BSA, ALB),
                    list(median = ~median(.x),
                         min = ~min(.x), max = ~max(.x)))
    ) |>
    tidyr::pivot_longer(dplyr::everything(),
                        names_to = c("Covariate", ".value"),
                        names_sep = "_"),
  digits = 2,
  caption = "Simulated cohort covariates (compare Table 1: weight 38.5 [12.9-79.9], height 149 [95-193], BSA 1.3 [0.58-2.1], albumin 34 [24-42])."
)
```

# Regimen comparison

Both regimens are given to the *same* virtual subjects, so the comparison is
paired. `rxSetSeed()` is called immediately before each arm's solve so the
two arms draw common random numbers.

Because `ss = 1` cannot converge on a model whose terminal half-life is 7 to
12 times the dosing interval, steady state is reached explicitly: `n_int`
twice-daily doses, observed over two windows. The first window is day 9.5 to
10, matching the cohort's median post-transplant time of 9.5 days -- the
sampling context of the paper's Figure 1. The second is the final interval,
which the approach-to-steady-state section above shows is fully accumulated
and is the basis of Figure 6 and Table 3.

```{r regimen-sim}
n_int  <- 120                  # 120 twice-daily doses = 60 days
t_end  <- 12 * n_int
win_fig1 <- seq(228, 240, by = 0.05)          # day 9.5-10
win_ss   <- seq(t_end - 12, t_end, by = 0.05) # accumulated steady state

make_arm <- function(dose_col, label, id_offset) {
  subjects |>
    dplyr::mutate(id = id + id_offset,
                  dose = .data[[dose_col]],
                  treatment = label) |>
    dplyr::rowwise() |>
    dplyr::reframe(
      id = id, WT = WT, ALB = ALB, OCC = 1, treatment = treatment,
      weight_band = weight_band, dose = dose,
      dplyr::bind_rows(
        tibble::tibble(time = 0, evid = 1, cmt = "depot",
                       ii = 12, addl = n_int - 1),
        tibble::tibble(time = c(win_fig1, win_ss), evid = 0,
                       cmt = "central", ii = 0, addl = 0)
      )
    ) |>
    dplyr::mutate(amt = ifelse(evid == 1, dose, NA_real_)) |>
    dplyr::select(-dose)
}

events <- dplyr::bind_rows(
  make_arm("dose_licensed",  "Licensed dose",       0L),
  make_arm("dose_optimized", "Model-informed dose", 1000L)
)
# Test the frame directly: wrapping this in unique() would strip the very
# duplicates being asserted against, making the check unfalsifiable.
stopifnot(!anyDuplicated(events[, c("id", "time", "evid")]))

solve_arm <- function(label) {
  rxode2::rxSetSeed(4242)   # common random numbers across the two arms
  rxode2::rxSolve(
    mod,
    events = events |> dplyr::filter(treatment == label),
    keep = c("treatment", "weight_band", "WT", "ALB"),
    returnType = "data.frame"
  )
}

sim <- dplyr::bind_rows(solve_arm("Licensed dose"),
                        solve_arm("Model-informed dose"))
stopifnot(all(sim$Cc[!is.na(sim$Cc)] >= 0))

# The two arms must have drawn the same etas: the same subject appears in
# both, so per-subject clearance (which carries etalcl but no dose term) must
# match across arms.
cl_pair <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(subject = ifelse(id > 1000L, id - 1000L, id)) |>
  dplyr::distinct(subject, treatment, cl) |>
  tidyr::pivot_wider(names_from = treatment, values_from = cl)
stopifnot(max(abs(cl_pair$`Licensed dose` - cl_pair$`Model-informed dose`)) < 1e-8)
```

## Figure 1 -- concentration-time profiles

The paper's Figure 1 pools observed concentrations against time after dose,
sampled a median of 9.5 days after transplantation. The panel below is the
day 9.5 to 10 window of the licensed-dose arm, so the accumulation state
matches.

```{r figure-1}
fig1_dat <- sim |>
  dplyr::filter(treatment == "Licensed dose", !is.na(Cc),
                time >= 228, time <= 240) |>
  dplyr::mutate(tad = time - 228)

fig1_dat |>
  dplyr::group_by(time = tad) |>
  dplyr::summarise(Q05 = quantile(Cc, 0.05), Q50 = median(Cc),
                   Q95 = quantile(Cc, 0.95), .groups = "drop") |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.7) +
  coord_cartesian(xlim = c(0, 15), ylim = c(0, 46)) +
  labs(x = "Time after dose (h)", y = "MPA concentration (mg/L)",
       title = "MPA profile on the licensed dose, day 9.5-10 after transplant",
       caption = paste("Median with 5th-95th percentile band. Comparable to Figure 1 of Heida 2024,",
                       "whose observed concentrations span roughly 0-46 mg/L over 0-15 h",
                       "with the peak cloud at 1-2 h."))

# The paper's Figure 1 axes run to 46 mg/L and the dense early cloud sits at
# 1-2 h after dose. Check the simulated peak timing and the envelope against
# those axis-derived bounds rather than against any single subject.
peak_time <- fig1_dat |>
  dplyr::group_by(id) |>
  dplyr::slice_max(Cc, n = 1, with_ties = FALSE) |>
  dplyr::ungroup()

cat(sprintf("Median Tmax %.2f h (10th-90th percentile %.2f-%.2f)\n",
            median(peak_time$tad),
            quantile(peak_time$tad, 0.1), quantile(peak_time$tad, 0.9)))

# Tmax is set by ktr, which carries no random effect, so it varies only
# through weight; the bound is deliberately wide of the 0.9-1.3 h realised
# spread but still fails on a mis-transcribed ktr or a lost transit stage
# (shape 1 absorption would put Tmax below 0.6 h).
stopifnot(median(peak_time$tad) > 0.6, median(peak_time$tad) < 3)
```

## Figure 6 -- target attainment on the two regimens

The target is AUC0-12h of 30 to 60 mg*h/L. The paper reports four things: the
licensed dose puts the median AUC *above* the target range, the
model-informed dose puts it *inside*, the model-informed distribution is
*narrower*, and target attainment rises from *35% to 42%*.

## Deterministic comparison

These four claims are gated on a **typical-value** comparison rather than on
the between-subject-variability cohort, and the reason is reproducibility.
The covariates above are drawn with `set.seed()`, which seeds R's RNG and is
identical on every machine; the etas are drawn by rxode2, whose streams are
partitioned per solver thread and therefore differ between a 16-thread
workstation and a 2-core CI runner. Removing the etas leaves a comparison
that is fully deterministic everywhere, and at steady state it reduces
exactly to `dose / CL` per subject -- the identity already gated above. The
between-subject cohort is still simulated, and drives Figure 6 below; it is
simply not what the assertions hang on.

Measured across 1, 2, 4 and 16 solver threads, the with-variability cohort
gave licensed medians of 63.9 to 72.8 and an interquartile-range ratio of
0.70 to **1.02** -- i.e. the "narrower distribution" claim *inverts* on some
thread counts purely through resampling. That is precisely the kind of
assertion that passes where it is written and fails in CI, so it is reported
below but not gated on the stochastic cohort.

```{r regimen-typical}
typical_regimen <- subjects |>
  dplyr::mutate(
    cl = eq_params(WT, ALB)$cl,
    `Licensed dose`       = dose_licensed / cl,
    `Model-informed dose` = dose_optimized / cl
  ) |>
  tidyr::pivot_longer(c(`Licensed dose`, `Model-informed dose`),
                      names_to = "treatment", values_to = "auc")

typ_summary <- typical_regimen |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    `Median AUC`    = median(auc),
    `25th pct`      = quantile(auc, 0.25),
    `75th pct`      = quantile(auc, 0.75),
    `IQR width`     = IQR(auc),
    `In target (%)` = 100 * mean(auc >= 30 & auc <= 60),
    .groups = "drop"
  )

knitr::kable(typ_summary, digits = 1,
             caption = "Typical-value regimen comparison (no between-subject variability). Heida 2024 reports 35% target attainment on the licensed dose and 42% on the model-informed dose.")

tl <- typ_summary[typ_summary$treatment == "Licensed dose", ]
to <- typ_summary[typ_summary$treatment == "Model-informed dose", ]

# Deterministic given set.seed(): no rxode2 RNG is involved, so these are the
# assertions that carry the paper's four claims. Realised values are
# licensed median 72.3 / IQR 54.3 / 31.5% attainment and model-informed
# median 57.8 / IQR 38.6 / 40.0% attainment.
stopifnot(
  tl$`Median AUC` > 60,                                     # licensed median above target
  to$`Median AUC` > 30, to$`Median AUC` < 60,               # model-informed median inside target
  to$`IQR width` < 0.85 * tl$`IQR width`,                   # narrower distribution
  abs(tl$`In target (%)` - 35) < 6,                         # paper: 35% on the licensed dose
  abs(to$`In target (%)` - 42) < 6,                          # paper: 42% on the model-informed dose
  to$`In target (%)` > tl$`In target (%)` + 5
)
```

## Figure 6 -- the distribution

```{r figure-6}
ss_dat <- sim |>
  dplyr::filter(!is.na(Cc), time >= t_end - 12) |>
  dplyr::mutate(tad = time - (t_end - 12))

auc_by_subject <- ss_dat |>
  dplyr::group_by(id, treatment, weight_band) |>
  dplyr::summarise(auc = trapz(tad, Cc), .groups = "drop") |>
  dplyr::mutate(treatment = factor(treatment,
                                   levels = c("Licensed dose", "Model-informed dose")))

ggplot(auc_by_subject, aes(treatment, auc)) +
  geom_boxplot(outlier.size = 0.6, width = 0.5) +
  geom_hline(yintercept = c(30, 60), linetype = "dashed") +
  scale_y_log10() +
  labs(x = NULL, y = "AUC0-12h (mg*h/L)",
       title = "Predicted steady-state AUC0-12h by regimen",
       caption = "Replicates Figure 6 of Heida 2024. Dashed lines: 30-60 mg*h/L target range.")

attainment <- auc_by_subject |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    `Median AUC` = median(auc),
    `25th pct`   = quantile(auc, 0.25),
    `75th pct`   = quantile(auc, 0.75),
    `IQR width`  = quantile(auc, 0.75) - quantile(auc, 0.25),
    `In target (%)` = 100 * mean(auc >= 30 & auc <= 60),
    .groups = "drop"
  )

knitr::kable(attainment, digits = 1,
             caption = "With-variability cohort. Reported for comparison with Figure 6; the gated claims are in the deterministic table above.")

lic <- attainment[attainment$treatment == "Licensed dose", ]
opt <- attainment[attainment$treatment == "Model-informed dose", ]

# Only robust properties are asserted on the stochastic cohort. The
# licensed-minus-model-informed median gap is a PAIRED contrast on the same
# subjects and was 10.2 to 17.3 mg*h/L across 1, 2, 4 and 16 threads; the
# bound of 5 sits well below that and still goes red if the banded scheme
# stops lowering exposure relative to weight-blind mg/m2 dosing. The
# attainment gain ran 4.5 to 8.5 points over the same runs.
# Deliberately NOT asserted here: the IQR ordering, which inverted at 4
# threads (ratio 1.024) -- see the deterministic table above.
stopifnot(
  lic$`Median AUC` - opt$`Median AUC` > 5,
  opt$`In target (%)` - lic$`In target (%)` > 2
)
```

# PKNCA validation

NCA is run over the accumulated steady-state dosing interval, grouped by
regimen so results roll up the way the paper reports them. The interval is
re-based to 0-12 h so it reads as a conventional AUC0-tau; the concentration
at `tad = 0` is the steady-state trough, not a pre-dose zero, which is the
correct anchor for a steady-state interval.

```{r pknca}
sim_nca <- ss_dat |>
  dplyr::select(id, time = tad, Cc, treatment)

# Guarantee a time = 0 row per (id, treatment). The re-based grid already
# starts at 0, so this is defensive only; existing rows win.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |>
    dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

# One dose row per subject at the start of the re-based interval.
dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::mutate(time = 0) |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "mg/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "mg")

intervals <- data.frame(
  start = 0, end = 12,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, cav = TRUE, auclast = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))

nca_res$result |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "cmin", "cav", "auclast")) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(Median = median(PPORRES),
                   `5th pct` = quantile(PPORRES, 0.05),
                   `95th pct` = quantile(PPORRES, 0.95),
                   .groups = "drop") |>
  dplyr::mutate(Parameter = nlmixr2lib::ncaParamLabel(PPTESTCD), .after = treatment) |>
  dplyr::select(-PPTESTCD) |>
  dplyr::rename("Regimen" = treatment) |>
  knitr::kable(digits = 2,
               caption = "Steady-state NCA over the 12 h dosing interval, by regimen.")
```

## Comparison against the published dosing relation

The paper publishes no NCA table, so the reference side of this comparison is
its own `dose = clearance * AUC` relation -- the arithmetic Table 3 was built
from -- evaluated per weight band at the band's representative weight and the
33 g/L simulation albumin. The simulated side is a typical-value solve of the
packaged ODE system with the same covariates. Agreement therefore tests the
implemented structure, allometry, albumin term and dose units against the
published relation.

```{r nca-compare}
# Representative weight = the midpoint of each Table 3 band. Albumin is set
# to 33 g/L, the mean the paper used for its own dosing simulations (Methods,
# Development of an improved starting dose) -- NOT the 34 g/L normalisation
# reference of Eq. 1, which is the cohort median. The distinction matters:
# the albumin exponent is -2.49, so the 1 g/L difference moves clearance by
# 7.7% and the 40-59 kg band sits on either side of the 60 mg*h/L target
# boundary depending on which is used.
band_alb <- 33

bands <- tibble::tibble(
  treatment = c("<20 kg", "20-39 kg", "40-59 kg", "60-90 kg"),
  WT        = c(15, 30, 50, 75),
  dose      = c(250, 500, 750, 1000)
)

# Typical-value solve (no random effects) dosed to full accumulation, with
# the final interval re-based to 0-12 h.
solve_typical_ss <- function(WT, ALB, dose) {
  ev <- rxode2::et(amt = dose, cmt = "depot", ii = 12, addl = n_int - 1) |>
    rxode2::et(win_ss, cmt = "central")
  ev <- as.data.frame(ev)
  ev$WT <- WT; ev$ALB <- ALB; ev$OCC <- 1
  out <- rxode2::rxSolve(mod, ev, omega = NA, sigma = NA,
                         returnType = "data.frame")
  out <- out[!is.na(out$Cc), ]
  out$time <- out$time - (t_end - 12)
  out
}

typical_nca <- bands |>
  dplyr::rowwise() |>
  dplyr::reframe(
    treatment = treatment,
    solve_typical_ss(WT, band_alb, dose) |> dplyr::select(time, Cc)
  ) |>
  dplyr::mutate(id = 1L)

typ_conc <- PKNCA::PKNCAconc(as.data.frame(typical_nca),
                             Cc ~ time | treatment + id,
                             concu = "mg/L", timeu = "h")
typ_dose <- PKNCA::PKNCAdose(
  bands |> dplyr::mutate(id = 1L, time = 0) |>
    dplyr::select(id, time, amt = dose, treatment) |> as.data.frame(),
  amt ~ time | treatment + id, doseu = "mg"
)
typ_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  typ_conc, typ_dose,
  intervals = data.frame(start = 0, end = 12, auclast = TRUE, cmax = TRUE)
))

published <- bands |>
  dplyr::mutate(auclast = dose / eq_params(WT, band_alb)$cl) |>
  dplyr::select(treatment, auclast)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = typ_res,
  reference = published,
  by        = "treatment",
  units     = c(auclast = "mg*h/L"),
  tolerance_pct = 20
)

knitr::kable(cmp, caption = paste("Simulated steady-state AUC0-12h against the paper's",
                                  "dose / clearance relation, by Table 3 weight band.",
                                  "* differs from reference by >20%."),
             align = c("l", "l", "r", "r", "r"))

# Deterministic typical-value comparison of an exact identity: no random
# effects on either side, so this is asserted tightly.
pct <- as.numeric(gsub("[^0-9.-]", "", cmp$`% diff`))
stopifnot(all(abs(pct) < 0.5))

# And every band lands inside the published 30-60 mg*h/L target, which is the
# property Table 3 was constructed to have. Deterministic arithmetic on
# published values, so this is asserted exactly at the stated boundaries.
band_auc <- published$auclast
stopifnot(all(band_auc >= 30), all(band_auc <= 60))
knitr::kable(
  bands |> dplyr::mutate(`AUC0-12h (mg*h/L)` = band_auc) |>
    dplyr::rename("Weight band" = treatment, "Representative weight (kg)" = WT,
                  "Twice-daily dose (mg)" = dose),
  digits = 1,
  caption = paste("Table 3 dosing scheme evaluated at each band's representative weight",
                  "and the paper's 33 g/L simulation albumin; the 30-60 mg*h/L target is",
                  "met throughout.")
)
```

The margin is not large. At the 34 g/L normalisation reference instead of the
33 g/L simulation mean, the 40-59 kg band gives 60.3 mg*h/L -- just outside
the target. That is consistent with the paper's own hedge that "the optimized
dosing strategy still leads to some supratherapeutic exposure", and with its
stated preference for a slightly elevated AUC over a subtherapeutic one.

# Assumptions and deviations

## Errata and internal inconsistencies in the source

* **Albumin normalisation constant: 34 vs 35 g/L.** Methods states the
  covariate "was normalized by dividing the value by 35 g/L" and Eq. 1 prints
  `(albumin/35)^-2.5`, but Table 2 prints `(Albumin/34)^theta2` with
  `theta2 = -2.49` and the supplementary control stream computes
  `COVALB = (ALB/34)**THETA(6)` with `THETA(6) = -2.49`. The packaged model
  uses **34 g/L and -2.49**, because the executable control stream and the
  results table agree with each other, 34 g/L is the cohort median albumin
  (Table 1), and Eq. 1 additionally rounds the exponent to -2.5, marking it
  as a restatement of Table 2 rather than the fitted form.
* **Abstract vs Table 2 confidence intervals.** The abstract prints
  `Vc/F (24.9 L; 95% CI, 93.0-6.71E25)` and `Q/F (36.2 L h-1; 95% CI,
  9.63-74.7)`, whereas Table 2 prints `6.53-45.8` and `25.8-49.6`. The point
  estimates agree everywhere and only the Table 2 intervals are internally
  consistent (the abstract's Vc/F interval does not contain its own point
  estimate). Confidence intervals are not carried in the model file; this is
  recorded so a reader comparing the two is not misled.
* **Supplementary control stream `$PK` variable naming is off by one.** The
  block assigns `V2=THETA(2)` and `V3=THETA(3)` and sets `S2=V2`, but then
  forms `K30=CL/V3`, `K34=Q/V3` and `K43=Q/V4` -- and `V4` is never assigned.
  Read as compartment-numbered volumes (`V3` = volume of compartment 3, the
  central compartment; `V4` = compartment 4, peripheral), every remaining
  line is consistent, and that reading is confirmed independently by the
  `$THETA` comments (`24.9 ;V3`, `1590 ;V4`), by the `$OMEGA` comments
  (`2.42 ; 2 IIV V3`, matching the 320% CV that Table 2 assigns to Vc/F), and
  by Table 2 itself (Vc/F = 24.9 L, Vp/F = 1590 L). The packaged model
  implements that unambiguous physical structure; the `V2` / `S2` lines are
  a transcription slip in the supplement with no effect on the intended
  model.
* **Table 3 caption.** The caption reads "Doses are described as twice the
  daily dose in milligrams" while the column header reads "Twice daily dose
  (mg)". The tabulated value is the dose given twice daily, not twice the
  daily dose: only that reading puts every band inside the 30-60 mg*h/L
  target through `dose = clearance * AUC`, as the table above shows.
* **Inter-individual variability variances.** Table 2 reports variability as
  percent CV and Methods gives the transform `CV = sqrt(exp(omega^2) - 1)`.
  Back-transforming the Table 2 percentages gives 0.13889, 2.41955, 0.33997
  and 0.19267; the supplementary control stream `$OMEGA` block lists 0.139,
  2.42, 0.337 and 0.19. The first two agree exactly; the last two differ in
  the third decimal, in a direction that rounding cannot explain (0.337
  back-transforms to 63.3% against the printed 63.6%; 0.19 to 45.8% against
  the printed 46.1%). The packaged model uses the **Table 2** values, since
  that is the peer-reviewed result carrying confidence intervals and
  shrinkage and is what a reader checks a reproduction against. The
  difference is immaterial to any simulation.

## Assumptions made in this vignette

* **Virtual cohort.** The paper generated its dosing-simulation population
  with PopGen from the ICRP database, which is not reproducible here. Weight
  is instead drawn lognormally about the Table 1 median of 38.5 kg with
  `sdlog = 0.43` and truncated to the observed 12.9-79.9 kg range; the
  `sdlog` was chosen so the truncation limits sit near the tails rather than
  fitted to anything. Height is derived from weight by a log-log fit through
  the three paired Table 1 anchors, a pairing established by the body surface
  area check above. Serum albumin uses the paper's own simulation setting
  (mean 33 g/L, CV 21%), truncated to the observed 24-42 g/L range because
  the -2.49 albumin exponent is steep and the model is not supported outside
  the observed range.
* **Target attainment is compared on typical values, not on the
  variability cohort.** The paper's 35% and 42% come from a 1000-subject
  PopGen cohort with a different weight distribution (ages 3-18 from a
  representative European population) than the Table 1-matched cohort used
  here, so exact agreement was not expected; the typical-value comparison
  nonetheless lands at 31.5% and 40.0%. The gated assertions allow 6
  percentage points either side of the published values.
* **Why the gate is on typical values.** Covariates are drawn with
  `set.seed()`, which seeds R's RNG and is reproducible everywhere. Etas are
  drawn by rxode2, whose RNG streams are partitioned per solver thread, so a
  different thread count draws a different cohort and no seed can make the
  two agree. Rendering the with-variability cohort at 1, 2, 4 and 16 threads
  gave licensed median AUC0-12h of 63.9 to 72.8 mg*h/L and an
  interquartile-range ratio of 0.70 to 1.02 -- meaning the paper's "narrower
  distribution" claim inverts on at least one thread count purely through
  resampling. Gating that claim on the stochastic cohort would have shipped a
  vignette that passes locally and fails in CI. It is gated on the
  deterministic typical-value comparison instead, where the ratio is 0.71,
  and the stochastic cohort carries only paired, wide-margin assertions.
* **Steady state is reached explicitly, not analytically.** rxode2's `ss = 1`
  steady-state dosing does not converge on this model -- the terminal
  half-life is 7 to 12 times the 12 h dosing interval across the covariate
  space, and the solver reports "could not solve the system" over much of it.
  Every steady-state quantity here is therefore produced by simulating 120
  twice-daily doses (60 days) and reading the final interval, which the
  approach-to-steady-state section shows is fully accumulated. The
  `dose = clearance * AUC` gate uses the equivalent single-dose form of the
  same identity, `AUC0-inf = dose / CL`, which superposition makes exact for
  a linear model and which needs no accumulation at all.
* **Table 3 is a starting-dose scheme, but its arithmetic is a steady-state
  one.** AUC0-12h reaches 95% of its steady-state value only after several
  weeks (see above), a consequence of the published model's large `Vp/F`.
  Exposure in the first days after transplantation -- when the scheme is
  applied -- is therefore materially below the AUC the scheme targets. This
  is a property of the published model, not of this implementation, and the
  paper does not discuss it.
* **Occasion index.** `OCC = 1` throughout, so the inter-occasion effect on
  relative bioavailability contributes a single draw per subject rather than
  re-drawing across dosing occasions.
* **No parameter in this model was derived from anything other than the
  paper's text, tables, equations and supplementary control stream.** No
  figure digitisation, author correspondence or upstream model was used.
