---
title: "Intra-CSF small molecules, LeiCNS-PK3.1 rat CNS PBPK (Hirasawa 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Intra-CSF small molecules, LeiCNS-PK3.1 rat CNS PBPK (Hirasawa 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
suppressPackageStartupMessages({
  library(nlmixr2lib)
  library(PKNCA)
  library(rxode2)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})
```

# Model and source

* Citation: Hirasawa M, de Lange ECM. Revisiting Cerebrospinal Fluid Flow Direction and Rate in Physiologically Based Pharmacokinetic Model. *Pharmaceutics.* 2022;14(9):1764. doi:[10.3390/pharmaceutics14091764](https://doi.org/10.3390/pharmaceutics14091764).
* Upstream framework: Saleh MAA, Bloemberg JM, Elassaiss-Schaap J, de Lange ECM. *J Pharmacokinet Pharmacodyn.* 2021;48(5):725-741. doi:[10.1007/s10928-021-09768-7](https://doi.org/10.1007/s10928-021-09768-7) (reference [6] of the 2022 paper). Its Supplementary table 2 supplies the rat CNS physiology and its Supplementary equations supply `Daq`, `P0`, `Qp` and `Qt`.

Hirasawa 2022 revisits the cerebrospinal-fluid (CSF) part of the LeiCNS-PK3.0 CNS PBPK
model. LeiCNS-PK3.0 moves drug through the CSF by one unidirectional, constant flow
(2.2 uL/min in the rat) from the lateral ventricles (LV) to the third and fourth
ventricles (TFV), cisterna magna (CM) and subarachnoid space (SAS). That model
over-predicts the elimination from CM of seven of ten small molecules given into
the CSF. The new **LeiCNS-PK3.1** model (Figure 1B) has:

* **bidirectional, site-dependent CSF movement**: downward and upward rates between
  LV, TFV and CM (`venQCSF,D`, `venQCSF,U`) and between CM and SAS (`cisQCSF,D`,
  `cisQCSF,U`), with one up/down ratio;
* a **drug-specific transfer clearance from SAS to plasma** (`sasQCSF`);
* a **choroid-plexus microvessel compartment** (CPMV), separated from the whole-brain
  microvessels, which feeds both blood-CSF barriers (BCSFB);
* a **correction factor on paracellular permeability** (`CFPPA`).

The three drug-independent CSF parameters were estimated from [3H]sucrose (ICV) and
[14C]sucrose (intracisternal) data. They were then fixed, and `sasQCSF` and `CFPPA`
were estimated for nine further molecules (Table 2). For three molecules the authors
also explored structural extensions (Supplementary Figures S2 and S3).

Fifteen model files are packaged, one per parameterisation:

```{r model-list}
models <- tibble::tribble(
  ~model, ~molecule, ~role,
  "Hirasawa_2022_sucrose_rat_pbpk", "sucrose", "CSF parameters estimated (Table 2)",
  "Hirasawa_2022_inulin_rat_pbpk", "inulin", "Table 2",
  "Hirasawa_2022_morphine6Glucuronide_rat_pbpk", "morphine-6-glucuronide", "Table 2",
  "Hirasawa_2022_morphine_rat_pbpk", "morphine", "Table 2",
  "Hirasawa_2022_atenolol_rat_pbpk", "atenolol", "Table 2; van Bree 1989 plasma model",
  "Hirasawa_2022_atenolol_microdialysis_rat_pbpk", "atenolol", "Table 2; de Lange 1994 data, Saleh 2021 plasma model",
  "Hirasawa_2022_acetaminophen_rat_pbpk", "acetaminophen", "Table 2; van Bree 1989 plasma model",
  "Hirasawa_2022_acetaminophen_microdialysis_rat_pbpk", "acetaminophen", "Table 2; de Lange 1994 plasma model",
  "Hirasawa_2022_antipyrine_rat_pbpk", "antipyrine", "Table 2",
  "Hirasawa_2022_cefodizime_rat_pbpk", "cefodizime", "Table 2",
  "Hirasawa_2022_guanidinosuccinicAcid_rat_pbpk", "guanidinosuccinic acid", "Table 2",
  "Hirasawa_2022_ziconotide_rat_pbpk", "ziconotide", "Table 2",
  "Hirasawa_2022_morphine_bsc_rat_pbpk", "morphine", "Figure S2: + brain/spinal-cord surface compartment",
  "Hirasawa_2022_acetaminophen_periventricular_rat_pbpk", "acetaminophen", "Figure S3: + periventricular clearance",
  "Hirasawa_2022_antipyrine_periventricular_rat_pbpk", "antipyrine", "Figure S3: + periventricular clearance"
)
knitr::kable(models)
```

Atenolol and acetaminophen each have two files because the paper uses two plasma
PK models for them (Supplementary Table S1). One goes with the van Bree 1989 CSF
data, to which Table 2 was fitted (Figure 4C,E); the other goes with the de Lange
1994 microdialysis data (Figure 4D,F). The CNS parameters are the same in both.

## CNS structure

```
plasma <-(Q_CBF)-> brain_vascular <-(BBB)-> brain_ecf <-> brain_cell_membrane <-> brain_icf <-> brain_lysosome
plasma <-(Q_CPBF)-> brain_choroid_plexus_vascular <-(BCSFB)-> brain_csf_lv, brain_csf_tfv
brain_ecf --(Q_ECF)--> brain_csf_lv <-(venQ D/U)-> brain_csf_tfv <-(venQ D/U)-> brain_csf_cm <-(cisQ D/U)-> brain_csf_sas --(sasQCSF)--> plasma
```

All concentrations are unbound. The plasma model is the empirical unbound model of
Supplementary Table S1. It is mass-balanced with the CNS, which is what lets drug
dosed into the CSF reach plasma. Neutral drug crosses a barrier by both the
paracellular and the transcellular route; charged drug only by the paracellular
route. Asymmetry factors (AF) carry active transport and are back-solved from the
reported `Kp,uu` values (Section 2.9; Supplementary Equations).

# Population

All data are literature group means from healthy rats, digitised by the authors
(Table 1). No original animals were studied. Rats are taken as 250 g (Section 2.5)
when per-kg doses are converted. Doses into the CSF are split across the CSF
compartments by the dosing-volume assumptions of Supplementary Table S3, and are
given as instantaneous boluses (Section 2.5). Subject counts are those of the cited
source studies and are not restated in the paper.

# Source trace

```{r source-trace}
trace_tab <- tibble::tribble(
  ~Quantity, ~`Source location`,
  "venQCSF,D 0.251, cisQCSF,D 1.07 uL/min; U/D ratio 1.59", "Table 2 (sucrose); fixed for all other molecules (footnote a)",
  "sasQCSF, CFPPA per molecule", "Table 2",
  "Upward rates = U/D ratio x downward rates", "Section 2.8 and Figure 1B",
  "CPMV volume (4 mg x 196 uL/g) and blood flow (4 mg x 4 mL/min/g)", "Section 2.7",
  "Plasma CLcen, CLcen-per1/2, Vcen, Vper1/2; IIV; RUV", "Supplementary Table S1",
  "MW, pKa, pKb, logP, fu,plasma, Kp,uu ECF / LV / CM", "Supplementary Table S2 (morphine and M6G row labels swapped; see Assumptions)",
  "Dose split across LV / TFV / CM / SAS", "Supplementary Table S3",
  "Doses and dosing volumes", "Table 1",
  "BSC variant: Qex 100 mL/min (FIX), sasQCSF,abs 1.22 uL/min (FIX), V_BSC 5.24 mL, CL_BSC 36.4 uL/min, CFPPA 0.0328", "Supplementary Figure S2B",
  "PV variant: sasQCSF and CL_PV; CFPPA = 1", "Supplementary Figure S3B",
  "Rat volumes, flows, surface areas, widths, pH", "Saleh 2021 Supplementary table 2",
  "Daq = 10^(-4.113 - 0.4609 log10 MW) x 60; P0 = 10^(0.939 logP - 6.21) x 60", "Saleh 2021 Supplementary equations (x 60 = per-second to per-minute)",
  "Qp = CFPPA x (Daq / width) x SA x f_para; Qt = 0.5 x P0 x SA x f_trans", "Saleh 2021 Supplementary equations; CFPPA from Section 2.6/2.8",
  "PHF = acidic x basic neutral fraction", "Saleh 2021 main text",
  "CL_wo = P0 x SA_BCM, CL_ow = CL_wo / 10^logP, Q_LYSO = P0 x SA_LYSO", "LeiCNS-PK3.0 as packaged in Saleh_2023_*_mouse_pbpk (see that vignette)",
  "AF selection rule (Kp,uu < 1: efflux AF, influx = 1; else vice versa)", "Section 2.9",
  "AF closed forms for the BBB, LV and TFV barriers", "Hirasawa 2022 Supplementary Equations"
)
knitr::kable(trace_tab, caption = "Source trace for the model equations and parameters.")
```

# Simulation helpers

```{r helpers}
BW_KG <- 0.25

typical_model <- function(name) {
  mod <- rxode2::rxode2(nlmixr2lib::readModelDb(name))
  if (any(!is.na(mod$iniDf$neta1))) mod <- rxode2::zeroRe(mod)
  mod
}

# Intra-CSF bolus split across the CSF compartments by Supplementary Table S3
# (volumes in uL reaching each compartment at the end of administration).
csf_dose <- function(amt, lv = 0, tfv = 0, cm = 0, sas = 0) {
  tot <- lv + tfv + cm + sas
  ev <- rxode2::et()
  for (x in list(c("brain_csf_lv", lv), c("brain_csf_tfv", tfv),
                 c("brain_csf_cm", cm), c("brain_csf_sas", sas))) {
    v <- as.numeric(x[2])
    if (v > 0) ev <- rxode2::et(ev, amt = amt * v / tot, cmt = x[1])
  }
  ev
}

solve_typical <- function(name, ev, times) {
  ev <- rxode2::et(ev, times)
  s <- suppressWarnings(rxode2::rxSolve(typical_model(name), ev,
                                        returnType = "data.frame"))
  s[order(s$time), ]
}
```

Every intra-CSF and IV scenario of Figures 3, 4, 5, S2 and S3:

```{r scenarios}
scen <- tibble::tribble(
  ~panel, ~model, ~route, ~amt, ~lv, ~tfv, ~cm, ~sas, ~tend,
  "3A [3H]sucrose ICV (kBq)", "Hirasawa_2022_sucrose_rat_pbpk", "csf", 5, 3.75, 7.5, 1.25, 0, 180,
  "3B [14C]sucrose IC (dpm)", "Hirasawa_2022_sucrose_rat_pbpk", "csf", 4 * 2.22e6, 0, 0, 17, 83, 2600,
  "3C inulin ICV", "Hirasawa_2022_inulin_rat_pbpk", "csf", 2.5e6 * BW_KG, 0, 3.75, 1.25, 0, 1440,
  "3D [14C]inulin IC (dpm)", "Hirasawa_2022_inulin_rat_pbpk", "csf", 2 * 2.22e6, 0, 0, 17, 83, 1440,
  "4A M6G ICV", "Hirasawa_2022_morphine6Glucuronide_rat_pbpk", "csf", 50 * 461.5, 3.75, 7.5, 1.25, 0, 170,
  "4B morphine ICV", "Hirasawa_2022_morphine_rat_pbpk", "csf", 50 * 285.3, 3.75, 7.5, 1.25, 0, 170,
  "4C atenolol ICV", "Hirasawa_2022_atenolol_rat_pbpk", "csf", 12000, 3.75, 6.25, 0, 0, 250,
  "4D atenolol ICV (microdialysis)", "Hirasawa_2022_atenolol_microdialysis_rat_pbpk", "csf", 150000, 3.75, 7.5, 3.75, 0, 125,
  "4E acetaminophen ICV", "Hirasawa_2022_acetaminophen_rat_pbpk", "csf", 210000, 3.75, 6.25, 0, 0, 125,
  "4F acetaminophen ICV (microdialysis)", "Hirasawa_2022_acetaminophen_microdialysis_rat_pbpk", "csf", 210000, 3.75, 6.25, 0, 0, 125,
  "4G antipyrine ICV", "Hirasawa_2022_antipyrine_rat_pbpk", "csf", 5e6, 3.75, 6.25, 0, 0, 300,
  "4H cefodizime ICV (nCi)", "Hirasawa_2022_cefodizime_rat_pbpk", "csf", 74, 3.75, 6.25, 0, 0, 125,
  "4I GSA ICV", "Hirasawa_2022_guanidinosuccinicAcid_rat_pbpk", "csf", 0.05 * BW_KG * 175.14 * 1000, 3.75, 6.25, 0, 0, 60,
  "4J ziconotide IT", "Hirasawa_2022_ziconotide_rat_pbpk", "csf", 100 * BW_KG * 1000, 0, 0, 0, 100, 360,
  "S2 morphine ICV, + BSC", "Hirasawa_2022_morphine_bsc_rat_pbpk", "csf", 50 * 285.3, 3.75, 7.5, 1.25, 0, 170,
  "S3C acetaminophen ICV, + PV", "Hirasawa_2022_acetaminophen_periventricular_rat_pbpk", "csf", 210000, 3.75, 6.25, 0, 0, 125,
  "S3D antipyrine ICV, + PV", "Hirasawa_2022_antipyrine_periventricular_rat_pbpk", "csf", 5e6, 3.75, 6.25, 0, 0, 300,
  "5A atenolol IV", "Hirasawa_2022_atenolol_rat_pbpk", "iv", 1e6, 0, 0, 0, 0, 250,
  "5B acetaminophen IV", "Hirasawa_2022_acetaminophen_rat_pbpk", "iv", 825000, 0, 0, 0, 0, 125,
  "5C antipyrine IV", "Hirasawa_2022_antipyrine_rat_pbpk", "iv", 2e7, 0, 0, 0, 0, 300,
  "5D cefodizime IV (nCi)", "Hirasawa_2022_cefodizime_rat_pbpk", "iv", 100 * BW_KG * 1000, 0, 0, 0, 0, 125,
  "5E ziconotide IV", "Hirasawa_2022_ziconotide_rat_pbpk", "iv", 100 * BW_KG * 1000, 0, 0, 0, 0, 360
)

sims <- lapply(seq_len(nrow(scen)), function(i) {
  r <- scen[i, ]
  ev <- if (r$route == "iv") {
    # Table 1: IV doses over 60 s where reported.
    rxode2::et(amt = r$amt, dur = 1, cmt = "central")
  } else {
    csf_dose(r$amt, r$lv, r$tfv, r$cm, r$sas)
  }
  times <- sort(unique(c(seq(0, r$tend, length.out = 241), 0.05, 0.5, 1, 2)))
  s <- solve_typical(r$model, ev, times)
  tibble::tibble(panel = r$panel, time = s$time, plasma = s$Cc,
                 ECF = s$Cbrain_ecf, CM = s$Cbrain_csf_cm, SAS = s$Cbrain_csf_sas)
}) |>
  bind_rows() |>
  mutate(panel = factor(panel, levels = scen$panel))
```

# Validation

The paper's estimated parameters come from a fit to digitised literature data, and
no NCA results are reported, so the model is gated against what the paper states
exactly: (1) the closed-form asymmetry factors of the Supplementary Equations,
(2) the `Kp,uu` values that define them, and (3) the model curves of the figures.

## Gate 1: asymmetry factors against the printed closed forms

The model files compute the asymmetry factors by a short sequential solution of
the steady-state CNS equations. Below, the Supplementary Equations are transcribed
verbatim and evaluated with each model's own parameters. The two must agree to
machine precision.

```{r gate1}
# Supplementary Equations, Hirasawa 2022 (symbols as printed).
af_bbb_ef <- function(QCBF, QE, KE, pB, tB, PM, PE, AFin)
  -(((QCBF + AFin * tB * PM + pB) * QE + pB * QCBF) * KE - (AFin * tB * PM + pB) * QCBF) /
    (tB * PE * QCBF * KE)
af_bbb_in <- function(QCBF, QE, KE, pB, tB, PM, PE, AFef)
  -(((QCBF + pB) * QE + (AFef * tB * PE + pB) * QCBF) * KE - pB * QCBF) /
    (tB * PM * QE * KE - tB * PM * QCBF)
af_lv_ef <- function(Q, QE, KE, KLV, KCM, vD, vU, cD, cU, s, pL, tL, PM, PL, AFin)
  (Q * KCM * vU * ((s + cU) * vU + cD * s) - (s + cU) * Q * KLV * vD^2 +
     ((Q * QE * KE - pL * Q * KLV + (AFin * tL * PM + pL) * (QE * KE - cD * KCM + Q)) * s -
        pL * Q * cU * KLV + Q * QE * cU * KE + (AFin * tL * PM + pL) * (QE * cU * KE + Q * cU)) * vD) /
  (vD * tL * PL * Q * KLV * (s + cU))
af_lv_in <- function(Q, QE, KE, KLV, KCM, vD, vU, cD, cU, s, pL, tL, PM, PL, AFef)
  -(Q * KCM * vU * ((s + cU) * vU + cD * s) - (s + cU) * Q * KLV * vD^2 +
      (((Q + pL) * QE * KE - pL * cD * KCM + pL * Q) * s - (AFef * tL * PL + pL) * Q * KLV * (s + cU) +
         cU * ((Q + pL) * QE * KE + pL * Q)) * vD) /
  (((QE * KE - cD * KCM + Q) * s + QE * cU * KE + Q * cU) * tL * PM * vD)
af_tfv_ef <- function(Q, QE, KE, KLV, KCM, vD, vU, cD, cU, s, pT, tT, PM, PT, AFin)
  -(Q * KCM * vU * ((s + cU) * vU + ((cD + pT) * s + pT * cU)) - (s + cU) * Q * KLV * vD^2 +
      ((Q * cD * KCM + (AFin * tT * PM + pT) * (cD * KCM - QE * KE - Q)) * s -
         (AFin * tT * PM + pT) * cU * (QE * KE + Q)) * vD + pT * Q * cD * KCM * s) /
  (((s + cU) * vU + cD * s) * tT * PT * Q * KCM)
af_tfv_in <- function(Q, QE, KE, KLV, KCM, vD, vU, cD, cU, s, pT, tT, PM, PT, AFef)
  (Q * KCM * vU * ((s + cU) * vU + cD * s + (AFef * tT * PT + pT) * (s + cU)) -
     (s + cU) * Q * KLV * vD^2 + Q * cD * KCM * s * vD -
     ((QE * KE - cD * KCM + Q) * s + QE * cU * KE + Q * cU) * pT * vD +
     (AFef * tT * PT + pT) * Q * cD * KCM * s) /
  (vD * tT * PM * (s * (QE * KE - cD * KCM + Q) + QE * cU * KE + Q * cU))

af_models <- models$model[!models$molecule %in% c("sucrose", "inulin")]
gate1 <- lapply(af_models, function(nm) {
  # Derived quantities come from the solve; ini() constants from the model.
  s <- c(as.list(solve_typical(nm, rxode2::et(amt = 1, cmt = "central"), 1)[1, ]),
         as.list(typical_model(nm)$theta))
  sas <- if (!is.null(s$Q_SAS)) s$Q_SAS else s$Q_SAS_abs
  a <- list(s$Q_CPBF, s$Q_ECF, s$kpuu_ecf, s$kpuu_lv, s$kpuu_cm, s$Q_CSF_ven_D,
            s$Q_CSF_ven_U, s$Q_CSF_cis_D, s$Q_CSF_cis_U, sas, s$Qp_BCSFB, s$Qt_BCSFB,
            s$PHF_MV, s$PHF_CSF, 1)
  if (s$kpuu_ecf < 1) {
    bbb <- c(s$AF_BBB_ef, af_bbb_ef(s$Q_CBF, s$Q_ECF, s$kpuu_ecf, s$Qp_BBB, s$Qt_BBB, s$PHF_MV, s$PHF_ECF, 1))
  } else {
    bbb <- c(s$AF_BBB_in, af_bbb_in(s$Q_CBF, s$Q_ECF, s$kpuu_ecf, s$Qp_BBB, s$Qt_BBB, s$PHF_MV, s$PHF_ECF, 1))
  }
  if (s$kpuu_cm < 1) {
    lv <- c(s$AF_LV_ef, do.call(af_lv_ef, a))
    tfv <- c(s$AF_TFV_ef, do.call(af_tfv_ef, a))
  } else {
    lv <- c(s$AF_LV_in, do.call(af_lv_in, a))
    tfv <- c(s$AF_TFV_in, do.call(af_tfv_in, a))
  }
  tibble::tibble(model = sub("^Hirasawa_2022_(.*)_rat_pbpk$", "\\1", nm),
                 `AF BBB` = bbb[1], `AF LV` = lv[1], `AF TFV` = tfv[1],
                 `max rel. diff vs printed` = max(abs(c(bbb[1] / bbb[2], lv[1] / lv[2], tfv[1] / tfv[2]) - 1)))
}) |>
  bind_rows()
knitr::kable(gate1, digits = 4, format.args = list(scientific = 3),
             caption = "Gate 1: model asymmetry factors (efflux if Kp,uu < 1, else influx) against the Supplementary Equations.")
```

```{r gate1-assert}
# Same parameters on both sides: the difference is pure floating-point error.
stopifnot(max(gate1$`max rel. diff vs printed`) < 1e-6)
```

Several factors are negative (for example the atenolol TFV efflux factor and the
guanidinosuccinic acid BBB efflux factor). This follows from the paper's rule of
back-solving one factor with the other fixed at 1. For example, with paracellular
permeability cut to 0.024% of its default (CFPPA = 0.000239), guanidinosuccinic
acid can only reach `Kp,uu,ECF` = 0.25 through a net transcellular influx. The
values are kept as the method produces them.

## Gate 2: steady-state Kp,uu recovery

`Kp,uu` is defined at steady state (Section 2.1). Under a long constant-rate
infusion into plasma, each Table 2 model must return the Supplementary Table S2
values it was parameterised with. This is an end-to-end test of the ODE topology,
the AF solution, the PHF block and the units. The three structural variants are
shown as well. Their factors are those of the base equations (see Assumptions), so
they do not reproduce the Table S2 values.

```{r gate2}
kpuu_tab <- tibble::tribble(
  ~molecule, ~kpuu_ecf, ~kpuu_cm,
  "morphine-6-glucuronide", 0.36, 0.029,
  "morphine", 0.37, 0.20,
  "atenolol", 0.042, 0.51,
  "acetaminophen", 0.22, 1.05,
  "antipyrine", 0.71, 1.51,
  "cefodizime", 0.19, 0.027,
  "guanidinosuccinic acid", 0.25, 0.25,
  "ziconotide", 0.018, 0.018
)
gate2 <- lapply(af_models, function(nm) {
  Tend <- 4e5
  ev <- rxode2::et(amt = 1000 * Tend, rate = 1000, cmt = "central")
  ev <- rxode2::et(ev, c(Tend - 1, Tend))
  s <- suppressWarnings(rxode2::rxSolve(typical_model(nm), ev, atol = 1e-12, rtol = 1e-10,
                                        maxsteps = 500000L, returnType = "data.frame"))
  z <- s[nrow(s), ]
  tibble::tibble(model = nm, `ECF sim` = z$Cbrain_ecf / z$Cc,
                 `LV sim` = z$Cbrain_csf_lv / z$Cc, `CM sim` = z$Cbrain_csf_cm / z$Cc)
}) |>
  bind_rows() |>
  left_join(models, by = "model") |>
  left_join(kpuu_tab, by = "molecule") |>
  mutate(variant = grepl("_bsc_|_periventricular_", model),
         `ECF err (%)` = 100 * (`ECF sim` / kpuu_ecf - 1),
         `CM err (%)` = 100 * (`CM sim` / kpuu_cm - 1),
         model = sub("^Hirasawa_2022_(.*)_rat_pbpk$", "\\1", model)) |>
  select(model, variant, kpuu_ecf, `ECF sim`, `ECF err (%)`, kpuu_cm, `LV sim`, `CM sim`, `CM err (%)`)
knitr::kable(gate2, digits = 4, caption = "Gate 2: steady-state unbound CNS-to-plasma ratios against Supplementary Table S2.")
```

```{r gate2-assert}
base <- gate2[!gate2$variant, ]
stopifnot(
  max(abs(base$`ECF err (%)`)) < 0.5,
  max(abs(base$`CM err (%)`)) < 0.5,
  max(abs(base$`LV sim` / base$kpuu_cm - 1)) < 0.005
)
```

## Gate 3: the AUC definition of Kp,uu, via PKNCA

For a linear system dosed into plasma, the ratio of `AUC0-inf` in a CNS compartment
to that in plasma equals the steady-state ratio. The paper uses this definition
when `Kp,uu` is not reported directly (Section 2.1). An IV bolus for each Table 2
model, with both AUCs computed by `PKNCA`, must return the Table S2 values.

```{r gate3}
grid <- sort(unique(c(0, exp(seq(log(0.01), log(4e6), length.out = 2500)))))
base_models <- af_models[!grepl("_bsc_|_periventricular_", af_models)]
auc_sims <- lapply(base_models, function(nm) {
  s <- suppressWarnings(rxode2::rxSolve(typical_model(nm),
                                        rxode2::et(rxode2::et(amt = 1000, cmt = "central"), grid),
                                        atol = 1e-14, rtol = 1e-11, maxsteps = 500000L,
                                        returnType = "data.frame"))
  tibble::tibble(model = nm, time = s$time, plasma = pmax(s$Cc, 0),
                 CM = pmax(s$Cbrain_csf_cm, 0), ECF = pmax(s$Cbrain_ecf, 0))
}) |>
  bind_rows()

conc_long <- auc_sims |>
  pivot_longer(c(plasma, CM, ECF), names_to = "analyte", values_to = "conc") |>
  filter(!is.na(conc))
dose_long <- tidyr::expand_grid(model = base_models, analyte = c("plasma", "CM", "ECF")) |>
  mutate(time = 0, dose = 1000)
o_conc <- PKNCA::PKNCAconc(as.data.frame(conc_long), conc ~ time | model + analyte)
o_dose <- PKNCA::PKNCAdose(as.data.frame(dose_long), dose ~ time | model + analyte)
nca <- suppressWarnings(PKNCA::pk.nca(PKNCA::PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(start = 0, end = Inf, auclast = TRUE, cmax = TRUE, tmax = TRUE)
)))
gate3 <- as.data.frame(nca) |>
  filter(PPTESTCD == "auclast") |>
  select(model, analyte, PPORRES) |>
  pivot_wider(names_from = analyte, values_from = PPORRES) |>
  left_join(models, by = "model") |>
  left_join(kpuu_tab, by = "molecule") |>
  mutate(`AUC ratio ECF` = ECF / plasma, `AUC ratio CM` = CM / plasma,
         `ECF err (%)` = 100 * (`AUC ratio ECF` / kpuu_ecf - 1),
         `CM err (%)` = 100 * (`AUC ratio CM` / kpuu_cm - 1),
         model = sub("^Hirasawa_2022_(.*)_rat_pbpk$", "\\1", model)) |>
  select(model, kpuu_ecf, `AUC ratio ECF`, `ECF err (%)`, kpuu_cm, `AUC ratio CM`, `CM err (%)`)
knitr::kable(gate3, digits = 4, caption = "Gate 3: PKNCA AUC ratios after an IV bolus against Supplementary Table S2.")
```

```{r gate3-assert}
# Both sides come from one deterministic solve; the residual is trapezoidal error.
stopifnot(max(abs(c(gate3$`ECF err (%)`, gate3$`CM err (%)`))) < 1)
```

## Gate 4: the published model curves

The maintainers read the dashed model-prediction lines of Figures 3, 4, 5, S2 and
S3 by eye at a few time points. Reading a log axis by eye is precise to perhaps
25%. Each simulated value must fall within a factor of 1.5 of the reading. The
plasma models of these panels have no or small IIV, so the published median of 200
simulations is close to the typical-value prediction compared here.

```{r gate4}
fig_read <- tibble::tribble(
  ~panel, ~output, ~time, ~read,
  "3A [3H]sucrose ICV (kBq)", "CM", 50, 5, "3A [3H]sucrose ICV (kBq)", "CM", 100, 2.6,
  "3A [3H]sucrose ICV (kBq)", "CM", 180, 1.0, "3A [3H]sucrose ICV (kBq)", "SAS", 30, 3.3,
  "3A [3H]sucrose ICV (kBq)", "SAS", 100, 1.7, "3A [3H]sucrose ICV (kBq)", "SAS", 180, 0.68,
  "3B [14C]sucrose IC (dpm)", "CM", 1000, 1.7e5, "3B [14C]sucrose IC (dpm)", "CM", 2600, 8.5e4,
  "3B [14C]sucrose IC (dpm)", "plasma", 300, 8e5, "3B [14C]sucrose IC (dpm)", "plasma", 2600, 3.3e5,
  "4A M6G ICV", "CM", 100, 1.5e4, "4A M6G ICV", "CM", 170, 6e3,
  "4A M6G ICV", "SAS", 100, 1e4, "4A M6G ICV", "SAS", 170, 4.5e3,
  "4B morphine ICV", "CM", 100, 1e3, "4B morphine ICV", "CM", 170, 80,
  "4B morphine ICV", "SAS", 100, 400, "4B morphine ICV", "SAS", 170, 40,
  "4C atenolol ICV", "CM", 50, 2e3, "4C atenolol ICV", "CM", 240, 70,
  "4C atenolol ICV", "plasma", 240, 0.6,
  "4D atenolol ICV (microdialysis)", "plasma", 125, 47, "4D atenolol ICV (microdialysis)", "ECF", 125, 2,
  "4E acetaminophen ICV", "CM", 120, 300, "4E acetaminophen ICV", "plasma", 120, 50,
  "4F acetaminophen ICV (microdialysis)", "plasma", 125, 7, "4F acetaminophen ICV (microdialysis)", "ECF", 125, 1.6,
  "4G antipyrine ICV", "CM", 200, 7e3, "4G antipyrine ICV", "CM", 300, 2e3,
  "4H cefodizime ICV (nCi)", "CM", 60, 50, "4H cefodizime ICV (nCi)", "CM", 120, 23,
  "4I GSA ICV", "CM", 20, 2.7e3, "4I GSA ICV", "CM", 60, 1.3e3,
  "4J ziconotide IT", "CM", 120, 50, "4J ziconotide IT", "CM", 360, 5,
  "4J ziconotide IT", "plasma", 360, 250,
  "S2 morphine ICV, + BSC", "CM", 100, 300, "S2 morphine ICV, + BSC", "CM", 170, 150,
  "S2 morphine ICV, + BSC", "SAS", 100, 180, "S2 morphine ICV, + BSC", "SAS", 170, 120,
  "S3C acetaminophen ICV, + PV", "CM", 95, 1e3, "S3C acetaminophen ICV, + PV", "CM", 120, 350,
  "S3C acetaminophen ICV, + PV", "plasma", 120, 55,
  "S3D antipyrine ICV, + PV", "CM", 100, 2e4, "S3D antipyrine ICV, + PV", "CM", 300, 6e3,
  "5A atenolol IV", "CM", 250, 35, "5A atenolol IV", "plasma", 250, 45,
  "5B acetaminophen IV", "CM", 120, 220, "5B acetaminophen IV", "plasma", 120, 200
)
sim_long <- sims |> pivot_longer(c(plasma, ECF, CM, SAS), names_to = "output", values_to = "sim")
gate4 <- fig_read |>
  rowwise() |>
  mutate(sim = {
    d <- sim_long[sim_long$panel == panel & sim_long$output == output, ]
    approx(d$time, d$sim, xout = time)$y
  }) |>
  ungroup() |>
  mutate(ratio = sim / read)
knitr::kable(gate4, digits = 3, format.args = list(scientific = 3),
             caption = "Gate 4: typical-value simulation against values read from the published model curves.")
```

```{r gate4-assert}
# Deterministic typical-value solve against figure readings: the bound covers
# the by-eye reading precision of a log axis.
stopifnot(all(abs(log(gate4$ratio)) < log(1.5)))
```

# Replicating the published figures

## Figure 3: CSF bulk-flow markers

```{r fig3, fig.height = 5}
sims |>
  filter(grepl("^3", panel)) |>
  pivot_longer(c(plasma, CM, SAS), names_to = "output", values_to = "conc") |>
  filter(conc > 0) |>
  ggplot(aes(time, conc, colour = output)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  facet_wrap(~panel, scales = "free") +
  labs(x = "Time (min)", y = "Unbound concentration", colour = NULL,
       caption = "Replicates Figure 3 of Hirasawa 2022 (typical value).") +
  theme_bw()
```

## Figure 4: eight further molecules

```{r fig4, fig.height = 8}
sims |>
  filter(grepl("^4", panel)) |>
  pivot_longer(c(plasma, ECF, CM, SAS), names_to = "output", values_to = "conc") |>
  filter(conc > 1e-3) |>
  ggplot(aes(time, conc, colour = output)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  facet_wrap(~panel, scales = "free", ncol = 3) +
  labs(x = "Time (min)", y = "Unbound concentration (ng/mL; nCi/mL for cefodizime)",
       colour = NULL, caption = "Replicates Figure 4 of Hirasawa 2022 (typical value).") +
  theme_bw()
```

## Figure 5: IV administration

```{r fig5, fig.height = 5}
sims |>
  filter(grepl("^5", panel)) |>
  pivot_longer(c(plasma, CM), names_to = "output", values_to = "conc") |>
  filter(conc > 1e-3) |>
  ggplot(aes(time, conc, colour = output)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  facet_wrap(~panel, scales = "free") +
  labs(x = "Time (min)", y = "Unbound concentration", colour = NULL,
       caption = "Replicates the LeiCNS-PK3.1 columns of Figure 5 of Hirasawa 2022 (typical value).") +
  theme_bw()
```

## Supplementary Figures S2 and S3: structural extensions

```{r figs2s3}
sims |>
  filter(grepl("^S", panel)) |>
  pivot_longer(c(plasma, CM, SAS), names_to = "output", values_to = "conc") |>
  filter(conc > 1e-3) |>
  ggplot(aes(time, conc, colour = output)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  facet_wrap(~panel, scales = "free") +
  labs(x = "Time (min)", y = "Unbound concentration (ng/mL)", colour = NULL,
       caption = "Replicates Supplementary Figures S2C,D and S3C,D of Hirasawa 2022.") +
  theme_bw()
```

## Between-subject variability: Figure 4D,F

The 95% prediction bands of Figure 4D,F come only from the IIV and residual error
of the de Lange 1994 plasma models (Section 2.12). The CNS model has no
variability of its own.

```{r vpc}
rxode2::rxSetSeed(20220824)
vpc <- lapply(c("4D atenolol ICV (microdialysis)", "4F acetaminophen ICV (microdialysis)"), function(p) {
  r <- scen[scen$panel == p, ]
  ev <- rxode2::et(csf_dose(r$amt, r$lv, r$tfv, r$cm, r$sas), seq(0.5, r$tend, by = 2.5))
  s <- suppressWarnings(rxode2::rxSolve(rxode2::rxode2(nlmixr2lib::readModelDb(r$model)),
                                        ev, nSub = 200, returnType = "data.frame"))
  s |>
    group_by(time) |>
    summarise(plasma_lo = quantile(sim, 0.025), plasma_med = median(Cc),
              plasma_hi = quantile(sim, 0.975),
              ecf_lo = quantile(Cbrain_ecf, 0.025), ecf_med = median(Cbrain_ecf),
              ecf_hi = quantile(Cbrain_ecf, 0.975), .groups = "drop") |>
    mutate(panel = p)
}) |>
  bind_rows()

ggplot(vpc, aes(time)) +
  geom_ribbon(aes(ymin = pmax(plasma_lo, 1e-2), ymax = plasma_hi), fill = "darkgreen", alpha = 0.2) +
  geom_line(aes(y = plasma_med, colour = "plasma")) +
  geom_ribbon(aes(ymin = ecf_lo, ymax = ecf_hi), fill = "firebrick", alpha = 0.2) +
  geom_line(aes(y = ecf_med, colour = "ECF")) +
  scale_y_log10() +
  facet_wrap(~panel, scales = "free") +
  labs(x = "Time (min)", y = "Unbound concentration (ng/mL)", colour = NULL,
       caption = "Median and 95% prediction interval of 200 simulated rats (Figure 4D,F).") +
  theme_bw()
```

# NCA of the intra-CSF profiles

The paper reports no NCA, so there is nothing to compare against. For reference,
this is a `PKNCA` summary of the simulated cisterna-magna profiles after each
intra-CSF dose, by panel.

```{r nca-csf}
cm_long <- sims |>
  filter(grepl("^[34]", panel)) |>
  transmute(panel = as.character(panel), time, conc = pmax(CM, 0)) |>
  filter(!is.na(conc))
cm_dose <- scen |>
  filter(grepl("^[34]", panel)) |>
  transmute(panel, time = 0, dose = amt)
nca_cm <- suppressWarnings(PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(as.data.frame(cm_long), conc ~ time | panel),
  PKNCA::PKNCAdose(as.data.frame(cm_dose), dose ~ time | panel),
  intervals = data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE, auclast = TRUE)
)))
as.data.frame(nca_cm) |>
  select(panel, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  rename("Panel" = panel, "Cmax (CM)" = cmax, "Tmax (min)" = tmax, "AUClast (CM)" = auclast) |>
  knitr::kable(digits = 3, format.args = list(scientific = 3),
               caption = "Simulated cisterna-magna NCA after intra-CSF dosing (units as in each panel).")
```

# Assumptions and deviations

## Assumptions

* **ODE equations.** The paper prints no differential equations. The topology is
  Figure 1B, with the LeiCNS-PK3.0 parenchymal block as packaged in the Saleh 2023
  mouse models. The steady state of the resulting equations reproduces the
  printed asymmetry-factor closed forms exactly (Gate 1). Those closed forms
  encode every flux, so the topology is pinned. The typical-value curves of every
  published panel are reproduced (Gate 4).
* **Plasma coupling.** The empirical plasma model is mass-balanced with the CNS:
  plasma exchanges with the brain and choroid-plexus microvessels at `Q_CBF` and
  `Q_CPBF` and receives `sasQCSF` from SAS. The LeiCNS-PK1.0 NONMEM control streams
  (Yamamoto 2017) are coupled the same way. Without it, drug dosed into the CSF could
  not reach plasma (Figures 3B,D, 4C-G, 4J). All concentrations, including those in
  the microvessels, are unbound. The Table S1 models describe unbound plasma data,
  and the closed forms contain no unbound-fraction term.
* **CPMV.** The choroid-plexus microvessel volume and blood flow are not subtracted
  from the whole-brain microvessel volume and cerebral blood flow. The paper does
  not say, and the amounts are below 2% and 1% respectively.
* **CFPPA** multiplies the paracellular clearance at the BBB and at both BCSFBs
  (Section 2.6: "paracellular permeability (PPA) across the BBB and BCSFB").
* **BCSFB width.** The Saleh 2021 table merges the BCSFB width cell with the BBB
  row, so both widths are 0.5 um. The BCSFB surface area (25 cm^2) is split equally
  between the LV and TFV barriers (Saleh 2021 footnote 13).
* **IIV scale.** Supplementary Table S1 gives IIV as a percentage of an
  exponential model. It is read as a coefficient of variation, with
  omega^2 = log(1 + CV^2). This affects only the prediction bands of Figure 4D,F
  and of the morphine model.
* **Dosing.** Rats weigh 250 g. Intra-CSF doses are instantaneous and split across
  compartments by Supplementary Table S3; the injected volume is not added to the
  compartment volumes. IV doses in Figure 5 are 1-min infusions (Table 1: 60 s).
  The [14C] doses are converted at 2.22e6 dpm/uCi.
* **Units.** Sucrose is in kBq (the [3H] ICV dataset). The [14C] IC dataset is in
  dpm and simulates identically, because the model is linear. Cefodizime is in nCi;
  all other molecules are in ng.
* **Asymmetry factors in the structural variants.** The paper does not say how the
  factors were computed for the Figure S2 and S3 extensions. Including the added
  pathway in the steady state fails for the periventricular variants: no positive
  choroid-plexus concentration satisfies the reported `Kp,uu`. For the BSC variant
  it gives curves that miss Figure S2. Applying the Supplementary Equations as
  printed reproduces Figures S2C,D and S3C,D (Gate 4). For the BSC variant this
  means using `sasQCSF,abs` in place of `sasQCSF` and ignoring the surface
  compartment; for the PV variants it means ignoring the periventricular clearance.
  As a result, the variants' steady-state ratios differ from Table S2 (Gate 2).

## Source oddities, carried through as published

* **Supplementary Table S2 swaps the morphine and morphine-6-glucuronide row
  labels.** The row labelled "Morphine-6-glucuronide" has MW 285.3, a basic charge
  class and logP 0.87, which is morphine (C17H19NO3, MW 285.3). The row labelled
  "Morphine" has MW 461.5, zwitterion and a calculated logP of 0.13, which is M6G
  (C23H27NO9, MW 461.5). Each row's fu, `Kp,uu` values and citations are consistent
  with the corrected assignment: the morphine row cites Saleh 2021 (a morphine
  dataset), and the M6G row cites the M6G studies of references [68-70]. Whole rows
  are reassigned. With this assignment the M6G and morphine panels of Figure 4A,B
  are reproduced (Gate 4).
* `sasQCSF` for guanidinosuccinic acid is negative (-0.411 uL/min): a net influx
  from plasma into SAS (Table 2 and Discussion). It is kept as published.
* Table 2 marks the morphine estimates (poor SAS fit) and the acetaminophen and
  antipyrine estimates (poor CM fit) as unsatisfactory. Supplementary Figures S2
  and S3 are the authors' attempts to fix these fits, and are packaged as separate
  models.
* The acetaminophen van Bree 1989 plasma model has an unusually low central
  clearance (0.075 mL/min) and a 4.07 L peripheral volume (Table S1). It is kept as
  published.

## Not extracted

* The original LeiCNS-PK3.0 simulations of Supplementary Figure S1 and the
  unidirectional-flow sensitivity analysis of Figure 2 use the predecessor model
  with default settings. They are not new parameterisations from this paper.
