---
title: "Digoxin (Hirai 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Digoxin (Hirai 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Hirai_2022_digoxin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Hirai_2022_digoxin"))$description`
* Article: <https://doi.org/10.1186/s40360-022-00552-y> (open access)

```{r load}
mod <- readModelDb("Hirai_2022_digoxin")
ui <- rxode2::rxode(mod)
tv <- rxode2::zeroRe(mod)
```

# Population

Hirai 2022 analysed 3465 trough serum digoxin concentrations from 391
consecutive Japanese adults with atrial fibrillation (AF) and heart failure
(HF) who took oral digoxin at Tokyo Women's Medical University Hospital
between 2008 and 2016 (Methods; Table 1). All patients had ACC/AHA stage C or
D HF. The cohort was elderly (67 +/- 14 years), light (57 +/- 15 kg) and 38%
female, with a median Cockcroft-Gault creatinine clearance of 56.5
[40.7-75.6] mL/min and a mean LVEF of 39 +/- 14%. Most patients took
0.125 mg/day (73%); 13% took 0.25 mg/day and 10% took 0.0625 mg/day. 16% were
on amiodarone, 8% on diltiazem and 6% on verapamil.

Concentrations drawn at least 6 h after the last dose and at least 5 days
after the start of therapy were treated as steady-state troughs. Because only
troughs were available, the authors fixed the absorption rate constant to
1.0 1/h and the apparent volume of distribution to 6.0 L/kg from the
literature and estimated only the apparent oral clearance. The model was fitted
in Phoenix NLME 8.1.

The same information is available programmatically via `ui$population`.

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| One-compartment, first-order absorption | n/a | Methods, *Population pharmacokinetic model development*; Results |
| `cl <- exp(lcl + etalcl) * (crcl_capped / 60)^e_crcl_cl * (1 + e_amio_cl * CONMED_AMIO)` | n/a | Results final model: `CL/F = 6.2 x (CLcr/60)^0.41 x (1 - 0.24 x [if amiodarone])`; exponential IIV per Methods `CL/F = tv CL/F x exp(eta)` |
| `crcl_capped <- min(CRCL, 120)` | 120 mL/min | Methods: CLcr above 120 mL/min replaced with 120 mL/min |
| `vc <- exp(lvc) * WT` | n/a | Results final model: `Vd/F = 6.0 x Body weight` |
| `Cc <- 1000 * central / vc` | n/a | mg/L to ng/mL (Table 4 concentration units) |
| `lka` | `fixed(log(1.0))` | Table 3 `ka (fixed)` = 1.000 1/h |
| `lcl` | `log(6.209)` | Table 3 `CL/F` = 6.209 L/h (RSE 2.83%) |
| `lvc` | `fixed(log(6.0))` | Table 3 `Vd/F (fixed)` = 6.000 L/kg |
| `e_crcl_cl` | 0.409 | Table 3 `CLCR on CL/F` = 0.409 (RSE 9.49%) |
| `e_amio_cl` | -0.238 | Table 3 `Amiodarone on CL/F` = -0.238 (RSE 3.16%) |
| `etalcl` | 0.344^2 = 0.118336 | Table 3 `omega CL/F` = 34.4% (RSE 2.9%) |
| `propSd` | 0.366 | Table 3 `Multiplicative` = 36.6% (RSE 3.2%); Methods `Cobs = Cpred x (1 + eps)` |

The Results equation prints the thetas rounded to two digits (6.2, 0.41, 0.24);
the model uses the three-decimal values of Table 3.

# Validation

## 1. Typical clearance and steady-state trough

The paper's clearance equation is evaluated directly and compared with the
model's `cl` output, and the solved steady-state trough is compared with the
closed-form one-compartment first-order-absorption trough
`Ctrough = F D ka / (V (ka - k)) [exp(-k tau)/(1 - exp(-k tau)) - exp(-ka tau)/(1 - exp(-ka tau))]`.
Both sides use the same parameters, so the tolerances are numerical.

```{r typical}
grid <- expand.grid(
  CRCL = c(30, 60, 90, 150),
  CONMED_AMIO = c(0, 1),
  dose = c(0.0625, 0.125, 0.25)
) |>
  mutate(id = row_number(), WT = 57)

ev_tv <- bind_rows(
  grid |> transmute(id, time = 0, amt = dose, ii = 24, ss = 1, evid = 1,
                    cmt = "depot", WT, CRCL, CONMED_AMIO),
  grid |> transmute(id, time = 24, amt = NA, ii = NA, ss = NA, evid = 0,
                    cmt = "central", WT, CRCL, CONMED_AMIO)
) |>
  arrange(id, time, desc(evid))

sim_tv <- rxode2::rxSolve(
  tv, events = ev_tv, returnType = "data.frame",
  rtol = 1e-10, atol = 1e-12, ssRtol = 1e-10, ssAtol = 1e-12, maxsteps = 1e6
) |>
  select(id, cl, vc, Cc) |>
  left_join(grid, by = "id")

trough_closed <- function(dose, cl, v, ka = 1, tau = 24) {
  k <- cl / v
  1000 * dose * ka / (v * (ka - k)) *
    (exp(-k * tau) / (1 - exp(-k * tau)) - exp(-ka * tau) / (1 - exp(-ka * tau)))
}

chk_tv <- sim_tv |>
  mutate(
    cl_paper = 6.209 * (pmin(CRCL, 120) / 60)^0.409 * (1 - 0.238 * CONMED_AMIO),
    trough_closed = trough_closed(dose, cl_paper, 6 * WT)
  )

stopifnot(
  nrow(chk_tv) == nrow(grid),
  max(abs(chk_tv$cl / chk_tv$cl_paper - 1)) < 1e-12,
  # CRCL cap: 150 mL/min gives the same clearance as 120 mL/min.
  all(abs(chk_tv$cl[chk_tv$CRCL == 150] -
            6.209 * 2^0.409 * (1 - 0.238 * chk_tv$CONMED_AMIO[chk_tv$CRCL == 150])) < 1e-12),
  max(abs(chk_tv$Cc / chk_tv$trough_closed - 1)) < 1e-6
)

chk_tv |>
  filter(CRCL != 150) |>
  transmute(
    `Daily dose (mg)` = dose, `CLcr (mL/min)` = CRCL,
    Amiodarone = ifelse(CONMED_AMIO == 1, "yes", "no"),
    `CL/F (L/h)` = signif(cl, 4),
    `Typical trough (ng/mL)` = signif(Cc, 3)
  ) |>
  knitr::kable(caption = "Typical-value clearance and steady-state trough (57 kg).")
```

## 2. Reproduction of Table 4 (probability of a toxic-range trough)

Hirai 2022 Table 4 gives, from a 1000-subject Monte Carlo simulation of the
final model, the probability that a steady-state trough is at least 0.9 ng/mL
or 1.2 ng/mL, for three daily doses, three CLcr values and with or without
amiodarone. The paper does not state the body weight used; the cohort mean of
57 kg (Table 1) is assumed. Body weight only moves the trough through the
fixed volume, so it has a small effect.

Instead of a random cohort, the between-subject distribution of CL/F is
integrated with 200 stratified quantiles of `etalcl` per cell, supplied through
`params`, and the multiplicative residual error `Cobs = Cpred x (1 + eps)` is
integrated analytically: `P(Cobs >= c) = 1 - Phi((c / Cpred - 1) / propSd)`.
The result is the exact model probability to within quadrature error, and does
not depend on the random-number stream.

```{r table4}
omega_cl <- ui$omega["etalcl", "etalcl"]
prop_sd <- ui$theta[["propSd"]]
nq <- 200

table4 <- tibble::tribble(
  ~dose,  ~CRCL, ~CONMED_AMIO, ~pub_09, ~pub_12,
  0.25,   90, 0, 61.9, 43.0,
  0.25,   60, 0, 72.7, 55.7,
  0.25,   30, 0, 86.9, 76.3,
  0.125,  90, 0, 18.7,  7.0,
  0.125,  60, 0, 28.1, 12.4,
  0.125,  30, 0, 51.3, 31.1,
  0.0625, 90, 0,  0.9,  0.0,
  0.0625, 60, 0,  1.8,  0.5,
  0.0625, 30, 0, 11.2,  3.1,
  0.25,   90, 1, 79.0, 65.7,
  0.25,   60, 1, 86.4, 74.5,
  0.25,   30, 1, 95.0, 88.1,
  0.125,  90, 1, 37.1, 19.0,
  0.125,  60, 1, 51.6, 31.5,
  0.125,  30, 1, 71.4, 54.4,
  0.0625, 90, 1,  3.7,  0.7,
  0.0625, 60, 1,  8.2,  1.7,
  0.0625, 30, 1, 24.3, 10.6
)

subj4 <- table4 |>
  mutate(cell = row_number()) |>
  tidyr::crossing(q = seq_len(nq)) |>
  mutate(
    id = row_number(),
    WT = 57,
    etalcl = sqrt(omega_cl) * qnorm((q - 0.5) / nq)
  )

ev4 <- bind_rows(
  subj4 |> transmute(id, time = 0, amt = dose, ii = 24, ss = 1, evid = 1,
                     cmt = "depot", WT, CRCL, CONMED_AMIO),
  subj4 |> transmute(id, time = 24, amt = NA, ii = NA, ss = NA, evid = 0,
                     cmt = "central", WT, CRCL, CONMED_AMIO)
) |>
  arrange(id, time, desc(evid))

# The etas are supplied through `params`, so no omega sampling is wanted:
# the typical-value model is solved and rxode2 would otherwise warn that a
# multi-subject simulation carries no omega.
sim4 <- suppressWarnings(rxode2::rxSolve(
  tv, events = ev4, params = subj4 |> select(id, etalcl),
  returnType = "data.frame", maxsteps = 1e6
)) |>
  select(id, cl, Cc) |>
  left_join(subj4 |> select(id, cell, dose, CRCL, CONMED_AMIO, etalcl), by = "id")

# Guard: every subject's clearance must carry its own supplied eta. A harness
# that silently drops `params` etas would solve everyone at the typical value.
cl_expected <- with(sim4, 6.209 * (pmin(CRCL, 120) / 60)^0.409 *
                      (1 - 0.238 * CONMED_AMIO) * exp(etalcl))
stopifnot(nrow(sim4) == nrow(subj4), max(abs(sim4$cl / cl_expected - 1)) < 1e-8)

pta <- sim4 |>
  group_by(cell, dose, CRCL, CONMED_AMIO) |>
  summarise(
    n = n(),
    sim_09 = 100 * mean(1 - pnorm((0.9 / Cc - 1) / prop_sd)),
    sim_12 = 100 * mean(1 - pnorm((1.2 / Cc - 1) / prop_sd)),
    nores_09 = 100 * mean(Cc >= 0.9),
    nores_12 = 100 * mean(Cc >= 1.2),
    .groups = "drop"
  ) |>
  left_join(table4, by = c("dose", "CRCL", "CONMED_AMIO")) |>
  mutate(diff_09 = sim_09 - pub_09, diff_12 = sim_12 - pub_12)

diffs <- c(pta$diff_09, pta$diff_12)
diffs_nores <- c(pta$nores_09 - pta$pub_09, pta$nores_12 - pta$pub_12)
stopifnot(
  nrow(pta) == 18L,
  all(pta$n == nq),
  # The published values carry Monte Carlo error from 1000 draws (binomial
  # SE up to 1.6 percentage points at p = 0.5), and the simulation weight is
  # not stated. Measured: median |diff| 0.5, maximum 1.9 points. A
  # mis-transcribed clearance, exponent or amiodarone factor moves the
  # mid-range cells by 5-20 points.
  median(abs(diffs)) < 1.5,
  max(abs(diffs)) < 4,
  # Without residual error the reproduction is clearly worse.
  max(abs(diffs_nores)) > 3 * max(abs(diffs))
)

pta |>
  transmute(
    `Dose (mg/day)` = dose, `CLcr (mL/min)` = CRCL,
    Amiodarone = ifelse(CONMED_AMIO == 1, "yes", "no"),
    `>= 0.9 model` = round(sim_09, 1), `>= 0.9 paper` = pub_09,
    `>= 1.2 model` = round(sim_12, 1), `>= 1.2 paper` = pub_12
  ) |>
  knitr::kable(caption = "Replicates Table 4 of Hirai 2022: percentage of steady-state troughs at or above 0.9 and 1.2 ng/mL.")
```

The packaged model reproduces every cell of Table 4 to within
`r round(max(abs(diffs)), 1)` percentage points (median absolute difference
`r round(median(abs(diffs)), 2)`). The agreement also supports two readings
that the paper leaves implicit: the Monte Carlo simulation included the
multiplicative residual error (without it the predicted probabilities differ
from Table 4 by up to `r round(max(abs(diffs_nores)), 1)` points), and the residual error enters as a normal eps on
the `(1 + eps)` scale.

## 3. Distribution of simulated troughs (Figure 2)

Figure 2 of Hirai 2022 shows violin plots of the predicted troughs by dose and
CLcr, without and with amiodarone. A stochastic cohort of 100 subjects per
cell is simulated here with residual error (`sim`), the observable the
Monte Carlo simulation summarised.

```{r fig2}
rxode2::rxSetSeed(20220214)
cells <- table4 |> select(dose, CRCL, CONMED_AMIO) |> mutate(cell = row_number())
subj2 <- cells |>
  tidyr::crossing(k = seq_len(100)) |>
  mutate(id = row_number(), WT = 57)

ev2 <- bind_rows(
  subj2 |> transmute(id, time = 0, amt = dose, ii = 24, ss = 1, evid = 1,
                     cmt = "depot", WT, CRCL, CONMED_AMIO),
  subj2 |> transmute(id, time = 24, amt = NA, ii = NA, ss = NA, evid = 0,
                     cmt = "central", WT, CRCL, CONMED_AMIO)
) |>
  arrange(id, time, desc(evid))

sim2 <- rxode2::rxSolve(mod, events = ev2, returnType = "data.frame",
                        maxsteps = 1e6) |>
  select(id, Cc, sim) |>
  left_join(subj2 |> select(id, dose, CRCL, CONMED_AMIO), by = "id")

stopifnot(nrow(sim2) == nrow(subj2), all(is.finite(sim2$sim)))

ggplot(sim2 |> mutate(
  Dose = factor(paste(dose, "mg"), levels = paste(c(0.25, 0.125, 0.0625), "mg")),
  CLcr = factor(paste(CRCL, "mL/min"), levels = paste(c(90, 60, 30), "mL/min")),
  Amiodarone = ifelse(CONMED_AMIO == 1, "With amiodarone", "Without amiodarone")
), aes(CLcr, pmax(sim, 0.01), fill = Dose)) +
  geom_violin(position = position_dodge(0.9), draw_quantiles = c(0.25, 0.5, 0.75),
              scale = "width") +
  geom_hline(yintercept = c(0.9, 1.2), linetype = "dashed") +
  facet_wrap(~Amiodarone) +
  scale_y_log10() +
  labs(x = "Creatinine clearance", y = "Trough serum digoxin (ng/mL)",
       caption = "Replicates Figure 2 of Hirai 2022 (57 kg; residual error included).")
```

## 4. Steady-state NCA with PKNCA

The paper reports no NCA. PKNCA is used here to check that the model's
steady-state dosing-interval exposure matches its clearance: at steady state
`AUC0-24 = F x Dose / (CL/F)`, and `Cavg = AUC0-24 / 24`. A stochastic cohort
of 100 subjects per CLcr / amiodarone group is dosed 0.125 mg once daily to
steady state and sampled every 0.5 h over one interval.

```{r nca}
rxode2::rxSetSeed(20220215)
grp <- expand.grid(CRCL = c(30, 60, 90), CONMED_AMIO = c(0, 1)) |>
  mutate(treatment = paste0("CLcr ", CRCL, ifelse(CONMED_AMIO == 1, " + amiodarone", "")))
subj_nca <- grp |>
  tidyr::crossing(k = seq_len(100)) |>
  mutate(id = row_number(), WT = pmin(pmax(rnorm(n(), 57, 15), 35), 100))

ev_nca <- bind_rows(
  subj_nca |> transmute(id, time = 0, amt = 0.125, ii = 24, ss = 1, evid = 1,
                        cmt = "depot", WT, CRCL, CONMED_AMIO),
  subj_nca |> tidyr::crossing(time = seq(0, 24, by = 0.5)) |>
    transmute(id, time, amt = NA, ii = NA, ss = NA, evid = 0,
              cmt = "central", WT, CRCL, CONMED_AMIO)
) |>
  arrange(id, time, desc(evid))

sim_nca <- rxode2::rxSolve(mod, events = ev_nca, returnType = "data.frame",
                           maxsteps = 1e6) |>
  select(id, time, Cc, cl) |>
  left_join(subj_nca |> select(id, treatment), by = "id")

conc <- sim_nca |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)
dose_df <- subj_nca |> transmute(id, time = 0, amt = 0.125, treatment)

pk_conc <- PKNCA::PKNCAconc(conc, Cc ~ time | treatment + id)
pk_dose <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)
intervals <- data.frame(start = 0, end = 24, auclast = TRUE, cmax = TRUE,
                        tmax = TRUE, cmin = TRUE, cav = TRUE)
nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(pk_conc, pk_dose, intervals = intervals))
nca_res <- as.data.frame(nca$result)

auc_chk <- nca_res |>
  filter(PPTESTCD == "auclast") |>
  select(id, treatment, auc = PPORRES) |>
  left_join(sim_nca |> distinct(id, cl), by = "id") |>
  mutate(auc_theory = 1000 * 0.125 / cl, rel = auc / auc_theory - 1)

stopifnot(
  nrow(auc_chk) == nrow(subj_nca),
  # Same drawn parameters on both sides; the only difference is trapezoidal
  # error over a 0.5 h grid (measured below 0.2%).
  max(abs(auc_chk$rel)) < 0.01
)

summary(nca)
```

# Assumptions and deviations

* **Omega scale.** Table 3 prints `omega CL/F = 34.4%`. The model takes this as
  the SD of the exponential eta (variance `0.344^2 = 0.118`). The alternative
  CV reading, `log(1 + 0.344^2) = 0.112`, differs by 5.6% in variance. When
  Table 4 is re-derived with the method above, the SD reading reproduces it
  slightly better (RMSE 0.84 vs 0.94 percentage points at 57 kg), which is
  consistent with but does not prove the choice.
* **Simulation body weight.** Table 4 and Figure 2 do not state the weight used
  in the Monte Carlo simulation; the cohort mean of 57 kg (Table 1) is used.
  Weights of 50 and 65 kg reproduce Table 4 less well (RMSE 1.8 and 1.0
  points).
* **Trough time.** Troughs are evaluated 24 h after a once-daily dose at steady
  state. The paper's troughs were drawn at least 6 h after the last dose, so
  real sampling times varied.
* **CLcr cap.** The Methods cap of 120 mL/min is applied inside `model()`, so
  users supply the uncapped Cockcroft-Gault value in `CRCL`.
* **Bioavailability.** Only CL/F and Vd/F are identifiable; there is no `F` in
  the model and the dose is the administered oral amount.
* **Amiodarone dose.** The amiodarone effect is a single fractional decrease
  in CL/F; the paper did not model amiodarone dose.
* **Errata.** A EuropePMC search on 2026-09-30 found no correction notice for
  this article.
