---
title: "Datopotamab deruxtecan ADC + DXd coupled population PK model (Hong 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Datopotamab deruxtecan ADC + DXd coupled population PK model (Hong 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 4.5)
```

```{r setup}
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Hong Y, Peigne S, Pan Y, Friberg Hietala S, McLaughlin A, Tajima N,
  Uema D, Zebger-Gong H, Tang Z, Zhou D, Abutarif M, Garimella T. *Population
  Pharmacokinetic Analysis of Datopotamab Deruxtecan (Dato-DXd), a
  TROP2-Directed Antibody-Drug Conjugate, in Patients With Advanced Solid
  Tumors.* CPT Pharmacometrics Syst Pharmacol. 2025;14(12):2149-2160.
* Article: <https://doi.org/10.1002/psp4.70118> (PMID 41035281).

The model couples two submodels (paper Figure 1):

1. **Dato-DXd submodel** -- two-compartment with parallel linear and
   Michaelis-Menten elimination from the central compartment.
2. **DXd submodel** -- one-compartment with linear clearance. Its central
   compartment is fed at a rate equal to the *total* (linear + Michaelis-Menten)
   Dato-DXd elimination rate, scaled by the DXd-to-Dato-DXd molecular-weight
   ratio (493.5 / 150,000) and a time-and-cycle-dependent drug-to-antibody
   ratio DAR(tad, CYCLE).

```
DAR(tad, CYCLE) = 4 * (0.25 + 0.75 * exp(-beta * tad)) *
                  (1                if CYCLE = 1
                   Factor1 = 0.696  otherwise)            (Hong 2025 Eq. 13)

rate_DXd_in     = (CL_lin * C_DatoDXd + Vmax * C_DatoDXd / (Km + C_DatoDXd))
                  * DAR * (493.5 / 150000)               (Hong 2025 Methods)
```

`tad` is the time since the most recent dose; `CYCLE` is the integer
treatment-cycle number, set to 1 over the first 21 days post-first-dose and
incremented at each new Q3W dosing cycle.

Concentrations are in `mg/L` (= `ug/mL`) for both analytes inside the model;
DXd is conventionally reported in `ng/mL` and is rescaled outside the model
(`Cdxd_ng_mL = Cdxd * 1000`).

# Population

Pooled analysis dataset (Hong 2025 Methods + Table S1):

* **Studies:** Phase 1 TROPION-PanTumor01 (TP01), Phase 2 TL05, Phase 3 TL01.
* **Subjects:** 729 patients with advanced or metastatic NSCLC and breast
  cancer.
* **Observations:** 9036 Dato-DXd PK observations + 9012 DXd PK observations.
* **Body-weight range:** 37.0 - 156 kg analysis range; reference 66 kg.
* **Reference subjects:**
  * Dato-DXd: 66 kg male, age 62 yr, albumin 38 g/L, tumor size 66 mm,
    region != Japan (Hong 2025 Figure 4 caption).
  * DXd: 66 kg male, region US, albumin 38 g/L, AST 22 U/L, total bilirubin
    0.4 mg/dL (Hong 2025 Figure 4 caption).
* **Dose regimens:** Phase 1 included a 0.27 - 10 mg/kg dose-escalation; Phase
  2 / 3 used the labeled 6 mg/kg Q3W. The approved label (Datroway) caps the
  flat dose at 540 mg for body weight >= 90 kg.

```{r inspect-metadata}
mod <- rxode2::rxode(readModelDb("Hong_2025_datopotamab"))
str(mod$meta$population, max.level = 1)
```

The covariates retained in the final model were body weight (mechanistic, on
all four Dato-DXd structural parameters and DXd CL/Vc), albumin and age (on
Dato-DXd CL_lin), region Japan (on Dato-DXd CL_lin), female sex (on Dato-DXd
CL_lin and Vc; DXd Vc), tumor size (on Dato-DXd Vmax), albumin / AST / total
bilirubin (on DXd CL), and region Europe / Rest of World (on DXd CL).
ADA, race, creatinine clearance, drug product, and broad hepatic-impairment
classifications were tested and excluded.

# Source trace

Every `ini()` entry in `inst/modeldb/specificDrugs/Hong_2025_datopotamab.R`
carries an in-file comment pointing to Hong 2025 Table 1, Table 2, or
Equations 8-15. The table below collates them.

| Parameter (nlmixr2lib)  | Value          | Units    | Source                   |
|-------------------------|---------------:|----------|--------------------------|
| `lcl_adc` -> CL_lin     | 0.386          | L/day    | Table 1, Eq. 8           |
| `lvc_adc` -> Vc         | 3.06           | L        | Table 1, Eq. 9           |
| `lq_adc`  -> Q          | 0.422          | L/day    | Table 1                  |
| `lvp_adc` -> Vp         | 2.88           | L        | Table 1, Eq. 10          |
| `lvmax`   -> Vmax       | 8.41           | mg/day   | Table 1, Eq. 12 (8410 ug/day) |
| `lkm`     -> Km         | 4.49           | mg/L     | Table 1, Eq. 11 (4490 ng/mL)  |
| `e_wt_cl_adc`           | 0.75 (fixed)   | -        | Table 1                  |
| `e_wt_vc_adc`           | 0.415          | -        | Table 1                  |
| `e_wt_vp_adc`           | 0.311          | -        | Table 1                  |
| `e_age_cl_adc`          | -0.306         | -        | Table 1                  |
| `e_alb_cl_adc`          | -0.788         | -        | Table 1                  |
| `e_japan_cl_adc`        | -0.219         | -        | Table 1                  |
| `e_sexf_cl_adc`         | -0.263         | -        | Table 1                  |
| `e_sexf_vc_adc`         | -0.160         | -        | Table 1                  |
| `e_tumsz_vmax`          | 0.125          | -        | Table 1, Eq. 12          |
| `lcl_dxd` -> CL_DXd     | 63.84          | L/day    | Table 2, Eq. 14 (2.66 L/h * 24) |
| `lvc_dxd` -> Vc_DXd     | 25.1           | L        | Table 2, Eq. 15          |
| `lfactor1` -> Factor1   | 0.696          | -        | Table 2, Eq. 13          |
| `lbeta`   -> beta       | 0.259          | 1/day    | Table 2, Eq. 13          |
| `e_wt_cl_dxd`           | 0.298 (fixed)  | -        | Table 2                  |
| `e_wt_vc_dxd`           | 0.530 (fixed)  | -        | Table 2                  |
| `e_alb_cl_dxd`          | 0.343          | -        | Table 2                  |
| `e_ast_cl_dxd`          | -0.154         | -        | Table 2                  |
| `e_tbili_cl_dxd`        | -0.139         | -        | Table 2                  |
| `e_eu_cl_dxd`           | 0.240          | -        | Table 2                  |
| `e_row_cl_dxd`          | 0.196          | -        | Table 2                  |
| `e_sexf_vc_dxd`         | -0.185         | -        | Table 2                  |
| `CcpropSd`              | 0.121          | fraction | Table 1 (additive RUV log scale) |
| `CdxdpropSd`            | 0.283          | fraction | Table 2 (additive RUV log scale) |

Inter-individual variability (log-normal; omega^2 = log(CV^2 + 1)):

| Parameter | %CV | omega^2  | Source  |
|-----------|----:|---------:|---------|
| CL_lin    | 27.2| 0.071375 | Table 1 |
| Vc        | 14.5| 0.020807 | Table 1 |
| Q         | 31.1| 0.092325 | Table 1 |
| Vp        | 31.2| 0.092893 | Table 1 |
| Vmax      | 19.2| 0.036201 | Table 1 |
| CL_DXd    | 31.4| 0.094033 | Table 2 |
| Vc_DXd    | 36.3| 0.123782 | Table 2 |

# Virtual cohort

Match the reference-subject covariates and the labeled dosing regimen
(6 mg/kg Q3W, 540 mg flat-dose cap for body weight >= 90 kg):

```{r cohort}
set.seed(20260427L)

n_subj   <- 200L
cycle_dy <- 21
n_cycles <- 5L

wt    <- exp(rnorm(n_subj, mean = log(66), sd = 0.20))
wt    <- pmin(pmax(wt, 37), 156)
sexf  <- rbinom(n_subj, 1, 0.50)
age   <- pmax(20, pmin(85, rnorm(n_subj, mean = 62, sd = 11)))
alb   <- pmax(20, pmin(50, rnorm(n_subj, mean = 38, sd = 4)))
ast   <- pmax(8,  pmin(80, rnorm(n_subj, mean = 22, sd = 6)))
tbili <- pmax(0.1, pmin(2, rlnorm(n_subj, log(0.4), 0.4)))
tumsz <- pmax(10, pmin(200, rlnorm(n_subj, log(66), 0.5)))
region <- sample(c("US", "JP", "EU", "RoW"), n_subj, replace = TRUE,
                 prob = c(0.45, 0.20, 0.20, 0.15))

make_subject <- function(id, wt_kg, sexf_ind, age_y, alb_g_L, ast_U_L,
                         tbili_mg_dL, tumsz_mm, reg) {
  dose_mg <- min(6 * wt_kg, 540)   # 6 mg/kg Q3W with 540 mg flat-dose cap

  doses <- rxode2::et() |>
    rxode2::et(amt = dose_mg, cmt = "adc_central", ii = cycle_dy,
               addl = n_cycles - 1L, time = 0)
  obs <- rxode2::et(seq(0.05, n_cycles * cycle_dy, length.out = 420),
                    cmt = "Cc")
  ev <- rbind(doses, obs)
  ev$id           <- id
  ev$WT           <- wt_kg
  ev$AGE          <- age_y
  ev$SEXF         <- sexf_ind
  ev$ALB          <- alb_g_L
  ev$AST          <- ast_U_L
  ev$TBILI        <- tbili_mg_dL
  ev$TUMSZ        <- tumsz_mm
  ev$REGION_JAPAN  <- as.integer(reg == "JP")
  ev$REGION_EUROPE <- as.integer(reg == "EU")
  ev$REGION_ROW    <- as.integer(reg == "RoW")
  ev$CYCLE        <- pmax(1L, as.integer(floor(ev$time / cycle_dy) + 1L))
  ev$dose_mg      <- dose_mg
  ev
}

events <- purrr::map_dfr(seq_len(n_subj),
  ~ make_subject(.x, wt[.x], sexf[.x], age[.x], alb[.x], ast[.x],
                 tbili[.x], tumsz[.x], region[.x])) |>
  as.data.frame()

stopifnot(length(unique(events$id)) == n_subj)
```

# Typical-subject simulation

Single typical-subject curve (66 kg male, age 62, albumin 38 g/L, AST 22 U/L,
total bilirubin 0.4 mg/dL, tumor 66 mm, region US, 6 mg/kg Q3W, 5 cycles)
with all between-subject and residual variability turned off:

```{r typical-sim}
typ <- readModelDb("Hong_2025_datopotamab") |> rxode2::zeroRe()

typ_events <- rxode2::et() |>
  rxode2::et(amt = 6 * 66, cmt = "adc_central", ii = cycle_dy,
             addl = n_cycles - 1L, time = 0) |>
  rxode2::et(seq(0.05, n_cycles * cycle_dy, length.out = 1500), cmt = "Cc")

typ_events$WT            <- 66
typ_events$AGE           <- 62
typ_events$SEXF          <- 0
typ_events$ALB           <- 38
typ_events$AST           <- 22
typ_events$TBILI         <- 0.4
typ_events$TUMSZ         <- 66
typ_events$REGION_JAPAN  <- 0
typ_events$REGION_EUROPE <- 0
typ_events$REGION_ROW    <- 0
typ_events$CYCLE         <- pmax(1L, as.integer(floor(typ_events$time / cycle_dy) + 1L))

typ_sim <- rxode2::rxSolve(typ, typ_events, returnType = "data.frame") |>
  mutate(
    Cc_ug_mL    = Cc,                # mg/L = ug/mL already
    Cdxd_ng_mL  = Cdxd * 1000        # mg/L -> ng/mL
  )
```

```{r fig2-typical, fig.cap = "Typical-subject Dato-DXd (top, ug/mL) and DXd (bottom, ng/mL) concentration-time profiles for a 66 kg male reference patient receiving 6 mg/kg Q3W for 5 cycles. Replicates the median trend of Hong 2025 Figure 2 (upper / lower panels)."}
typ_sim |>
  tidyr::pivot_longer(c(Cc_ug_mL, Cdxd_ng_mL),
                      names_to = "analyte", values_to = "conc") |>
  mutate(analyte = recode(analyte,
                          Cc_ug_mL   = "Dato-DXd (ug/mL)",
                          Cdxd_ng_mL = "DXd (ng/mL)")) |>
  ggplot(aes(time, conc)) +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  facet_wrap(~ analyte, ncol = 1, scales = "free_y") +
  scale_y_log10() +
  labs(x = "Time (day)", y = "Concentration",
       caption = "Replicates the median trend of Hong 2025 Figure 2.")
```

```{r fig1-dar, fig.cap = "Internal DAR(t) trajectory across five cycles for the typical subject. DAR resets at each dose (4 in cycle 1; 4 * 0.696 = 2.78 in cycles 2 onwards) and decays toward DAR0 * 0.25 in cycle 1 (= 1) and DAR0 * 0.25 * 0.696 in later cycles (= 0.696)."}
typ_sim |>
  ggplot(aes(time, dar)) +
  geom_line(colour = "darkred", linewidth = 0.7) +
  labs(x = "Time (day)", y = "DAR(t)",
       caption = "Hong 2025 Eq. 13.")
```

# Linear vs nonlinear elimination across the dose range

The paper's Figure 3 visualizes the contributions of CL_lin and CL_nonlin to
total Dato-DXd clearance versus the typical-subject Dato-DXd plasma
concentration. We replicate that plot from the typical-value parameters:

```{r fig3-clearance-split, fig.cap = "Typical-subject Dato-DXd linear, nonlinear, and total clearance (left axis) and percentage nonlinear (right axis) versus Dato-DXd plasma concentration. Replicates the qualitative message of Hong 2025 Figure 3: CL_lin dominates above ~50 ug/mL while nonlinear MM clearance dominates at sub-therapeutic concentrations."}
cl_grid <- tibble(
  C_ug_mL    = exp(seq(log(0.1), log(500), length.out = 200)),
  CL_lin     = 0.386,
  rate_lin   = CL_lin * C_ug_mL,
  rate_mm    = 8.41 * C_ug_mL / (4.49 + C_ug_mL),
  CL_nonlin  = rate_mm / C_ug_mL,
  CL_total   = CL_lin + CL_nonlin,
  pct_nl     = 100 * CL_nonlin / CL_total
)

ggplot(cl_grid, aes(C_ug_mL, CL_total)) +
  geom_line(aes(colour = "Total"), linewidth = 0.8) +
  geom_line(aes(C_ug_mL, CL_lin,    colour = "Linear"), linewidth = 0.8) +
  geom_line(aes(C_ug_mL, CL_nonlin, colour = "Nonlinear (MM)"), linewidth = 0.8) +
  scale_x_log10() +
  labs(x = "Dato-DXd plasma concentration (ug/mL)",
       y  = "Apparent clearance (L/day)",
       colour = "Clearance",
       caption = "Replicates Hong 2025 Figure 3 (typical-subject parameters).")
```

# Population (VPC-style) simulation

Full population simulation with all between-subject variability and residual
error preserved.

```{r population-sim}
mod <- readModelDb("Hong_2025_datopotamab")
pop_sim <- rxode2::rxSolve(mod, events,
                           keep = c("WT", "SEXF", "CYCLE", "dose_mg"),
                           returnType = "data.frame") |>
  mutate(
    Cc_ug_mL   = Cc,
    Cdxd_ng_mL = Cdxd * 1000
  )
```

```{r fig2-vpc-adc, fig.cap = "VPC-style 5/50/95th-percentile ribbon of simulated Dato-DXd concentrations vs time across 200 virtual subjects on 6 mg/kg Q3W (540 mg flat-dose cap >= 90 kg) for 5 cycles. Reproduces the Dato-DXd panel of Hong 2025 Figure 2."}
pop_sim |>
  group_by(time) |>
  summarise(
    p05 = quantile(Cc_ug_mL, 0.05, na.rm = TRUE),
    p50 = quantile(Cc_ug_mL, 0.50, na.rm = TRUE),
    p95 = quantile(Cc_ug_mL, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "steelblue", alpha = 0.3) +
  geom_line(colour = "steelblue4", linewidth = 0.8) +
  scale_y_log10() +
  labs(x = "Time (day)", y = "Dato-DXd concentration (ug/mL)",
       title = "Dato-DXd VPC -- 6 mg/kg Q3W",
       caption = "Replicates the upper panel of Hong 2025 Figure 2.")
```

```{r fig2-vpc-dxd, fig.cap = "VPC-style 5/50/95th-percentile ribbon of simulated DXd concentrations vs time across the same 200 virtual subjects. Reproduces the DXd panel of Hong 2025 Figure 2; cycle-1 to later-cycle scaling reflects the DAR Factor1 = 0.696 step."}
pop_sim |>
  group_by(time) |>
  summarise(
    p05 = quantile(Cdxd_ng_mL, 0.05, na.rm = TRUE),
    p50 = quantile(Cdxd_ng_mL, 0.50, na.rm = TRUE),
    p95 = quantile(Cdxd_ng_mL, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "tomato", alpha = 0.3) +
  geom_line(colour = "tomato4", linewidth = 0.8) +
  scale_y_log10() +
  labs(x = "Time (day)", y = "DXd concentration (ng/mL)",
       title = "DXd VPC -- 6 mg/kg Q3W",
       caption = "Replicates the lower panel of Hong 2025 Figure 2.")
```

# PKNCA validation

Per-cycle NCA on the simulated typical-subject (no IIV) curves -- the paper
reports relative metrics (R_ac for AUC, Cmax, Ctrough) rather than absolute
single-cycle Cmax / AUC values, so the validation target is the cycle-1 to
cycle-3 *ratio* and the qualitative ordering between cycles.

```{r pknca-typical}
typ_nca_input <- typ_sim |>
  filter(time > 0) |>
  mutate(
    cycle         = pmax(1L, as.integer(floor((time - 1e-9) / cycle_dy) + 1L)),
    time_in_cycle = time - (cycle - 1) * cycle_dy
  ) |>
  filter(cycle <= n_cycles) |>
  mutate(treatment = "Dato-DXd 6 mg/kg Q3W") |>
  transmute(id = 1L, treatment, cycle = as.factor(cycle),
            time = time_in_cycle, Cc = Cc_ug_mL, Cdxd = Cdxd_ng_mL)

dose_typ <- tibble(
  id        = 1L,
  treatment = "Dato-DXd 6 mg/kg Q3W",
  cycle     = factor(seq_len(n_cycles)),
  time      = 0,
  amt       = 6 * 66
)

intervals_q3w <- data.frame(
  start   = 0,
  end     = cycle_dy,
  cmax    = TRUE,
  tmax    = TRUE,
  auclast = TRUE
)

adc_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(typ_nca_input, Cc ~ time | treatment + id/cycle),
  PKNCA::PKNCAdose(dose_typ,      amt ~ time | treatment + id/cycle),
  intervals = intervals_q3w
))
adc_nca_summary <- summary(adc_nca)

dxd_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(typ_nca_input, Cdxd ~ time | treatment + id/cycle),
  PKNCA::PKNCAdose(dose_typ,       amt ~ time | treatment + id/cycle),
  intervals = intervals_q3w
))
dxd_nca_summary <- summary(dxd_nca)

adc_nca_summary
dxd_nca_summary
```

## Cycle 1 to cycle 3 accumulation comparison

```{r rac-comparison}
adc_pp <- as.data.frame(adc_nca$result) |>
  filter(PPTESTCD %in% c("auclast", "cmax")) |>
  mutate(analyte = "Dato-DXd")
dxd_pp <- as.data.frame(dxd_nca$result) |>
  filter(PPTESTCD %in% c("auclast", "cmax")) |>
  mutate(analyte = "DXd")

rac_tbl <- bind_rows(adc_pp, dxd_pp) |>
  group_by(analyte, PPTESTCD, cycle) |>
  summarise(value = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = cycle, values_from = value,
              names_prefix = "cycle_") |>
  mutate(R_ac_3 = cycle_3 / cycle_1)

knitr::kable(
  rac_tbl,
  digits  = c(NA, NA, 3, 3, 3, 3, 3, 2),
  caption = "Cycle-1 vs cycle-3 NCA metrics for the typical subject. Hong 2025 Results report R_ac (AUC) < 1.2 and R_ac (Cmax) ~ 1 at 6 mg/kg Q3W. Differences > 20% would indicate a structural problem with the model and would be investigated rather than tuned away."
)
```

# Body-weight-based dose-cap simulation

Hong 2025 Figure 5 evaluates the 540 mg flat-dose cap above 90 kg by
simulating WT subgroups [37,46), [46,60), [60,80), [80,90), [90,156] kg.
We replicate the Cmax3 and AUC3 boxplots for cycle 3 (using the typical
covariate vector, no IIV):

```{r fig5-wt-dosing, fig.cap = "Cycle-3 Dato-DXd AUC (left) and Cmax (right) by weight subgroup using typical-value parameters. Subjects in the >= 90 kg group receive the 540 mg cap. Replicates the qualitative message of Hong 2025 Figure 5: cap-vs-mg/kg dosing yields broadly similar exposures across subgroups."}
wt_groups <- tibble(
  group    = factor(c("[37,46)", "[46,60)", "[60,80)", "[80,90)", "[90,156]"),
                    levels = c("[37,46)", "[46,60)", "[60,80)", "[80,90)",
                               "[90,156]")),
  wt_med   = c(42, 53, 70, 85, 110)
)

simulate_typ_wt <- function(wt_kg) {
  dose_mg <- if (wt_kg >= 90) 540 else 6 * wt_kg
  ev <- rxode2::et() |>
    rxode2::et(amt = dose_mg, cmt = "adc_central", ii = cycle_dy,
               addl = 2L, time = 0) |>            # 3 cycles total
    rxode2::et(seq(0.05, 3 * cycle_dy, length.out = 1000), cmt = "Cc")
  ev$WT            <- wt_kg
  ev$AGE           <- 62
  ev$SEXF          <- 0
  ev$ALB           <- 38
  ev$AST           <- 22
  ev$TBILI         <- 0.4
  ev$TUMSZ         <- 66
  ev$REGION_JAPAN  <- 0
  ev$REGION_EUROPE <- 0
  ev$REGION_ROW    <- 0
  ev$CYCLE         <- pmax(1L, as.integer(floor(ev$time / cycle_dy) + 1L))
  out <- rxode2::rxSolve(typ, ev, returnType = "data.frame") |>
    filter(time > 2 * cycle_dy & time <= 3 * cycle_dy) |>
    summarise(
      auc3  = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
      cmax3 = max(Cc, na.rm = TRUE)
    )
  out$wt_kg <- wt_kg
  out$dose  <- dose_mg
  out
}

wt_sim <- purrr::map_dfr(wt_groups$wt_med, simulate_typ_wt) |>
  left_join(wt_groups, by = c("wt_kg" = "wt_med"))

wt_sim |>
  pivot_longer(c(auc3, cmax3), names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric,
                         auc3  = "Cycle-3 AUC (ug*day/mL)",
                         cmax3 = "Cycle-3 Cmax (ug/mL)")) |>
  ggplot(aes(group, value)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  facet_wrap(~ metric, scales = "free_y") +
  labs(x = "Body-weight group (kg)", y = NULL,
       caption = "Replicates Hong 2025 Figure 5 (typical-value approximation; full IIV simulation produces the boxplots in the paper).")
```

# Assumptions and deviations

- **Coupled parent + payload in one model file.** The paper presents a single
  joint analysis with two interconnected sub-models (Dato-DXd two-compartment
  ADC, DXd one-compartment payload). We extract them together in
  `Hong_2025_datopotamab.R`, mirroring the Li 2017 brentuximab precedent.
- **Concentration units mg/L (= ug/mL) for both analytes.** Dato-DXd is
  conventionally reported in `ug/mL` and DXd in `ng/mL`. Inside the ODE, both
  are integrated in `mg/L`. The vignette converts DXd to `ng/mL` after solve.
- **CL_DXd unit conversion.** Hong 2025 Table 2 reports CL_DXd as
  2.66 L/h; the model file stores it as 2.66 * 24 = 63.84 L/day so that
  every clearance / rate constant in the ODE is on the day timescale (matching
  beta = 0.259 /day, the cycle-dosing interval, and Dato-DXd CL_lin
  = 0.386 L/day). The abstract reports CL_DXd as 2.66 L/day; this is a unit
  typo in the abstract -- a half-life of 6.5 days is implausible for the
  small-molecule payload DXd, whereas 2.66 L/h gives a half-life of ~6.5 h
  consistent with the published exatecan congener. The Table 2 / Eq. 14 unit
  (L/h) is treated as authoritative.
- **AST unit clarification.** The Methods text reports the AST reference
  value as `22 g/L`; this is implausible for a transaminase activity (normal
  range ~10-40 U/L; 22 g/L would be a higher concentration than total
  serum albumin). The reference numeric value 22 is consistent with the
  canonical clinical-PK unit U/L (= IU/L), and that is what we use. The
  parameter point estimate `e_ast_cl_dxd = -0.154` is unaffected by the unit
  label.
- **No IIV on Dato-DXd Q (CL-Vc covariance).** Hong 2025 Tables 1 / 2 do not
  report a correlation block for any pair of structural parameters; every
  IIV term is implemented as a diagonal omega.
- **No eta on the residual error (RUV).** Paper Eq. 3 places an exponential
  IIV on the RUV magnitude with %CV reported as `IIV RUV` in Tables 1 and 2.
  nlmixr2's residual-error model does not accept a subject-level random
  effect on the SD, so this term is omitted from the implementation; the
  reported additive-on-log-scale SD (which equals proportional-in-linear-
  space SD) is used as the typical-subject residual SD. This will under-
  reproduce the tails of the residual distribution but match the median.
- **Region-grouping discrepancy between Dato-DXd and DXd CL covariates.**
  Hong 2025 evaluates Dato-DXd CL with reference = "any region != Japan"
  (so REGION_JAPAN = 1 implies the -0.219 fractional shift) but DXd CL with
  reference = "US or Japan" (so REGION_JAPAN does not enter the DXd CL
  expression at all -- Japan is grouped with US for the DXd reference).
  The model file uses the three indicators REGION_JAPAN, REGION_EUROPE,
  REGION_ROW with US as the implicit reference (all three = 0); REGION_JAPAN
  enters the Dato-DXd CL term only and not the DXd CL term, faithfully
  reproducing the paper's parameterization.
- **Cycle covariate.** `CYCLE` is a 1-based integer treatment cycle counter
  (1 in days [0, 21), 2 in [21, 42), and so on for Q3W dosing). The DAR
  equation reads `factor_dar = 1` in cycle 1 and `factor_dar = 0.696`
  thereafter, implemented with an `if` test so the simulation user must
  populate `CYCLE >= 1` for every observation row.
- **`tad()` tracks time since the most recent dose.** Within the first
  21 days post-first-dose, `tad()` increases monotonically from 0; at the
  next dose it resets to 0 again. The DAR equation therefore decays during
  each cycle and reverts to 4 (cycle 1) or 4 * 0.696 = 2.78 (cycles 2+) at
  each new dose.
- **No bioavailability term.** Dato-DXd is administered IV; doses are
  written directly into `adc_central` and there is no `depot` compartment.
- **Population metadata.** Median age, median weight, sex distribution, and
  per-region patient counts are not enumerated in the publicly-available
  trimmed text used during extraction; the `population` block reports the
  pooled-cohort summary directly cited (n = 729, 3 studies, 9036 + 9012
  observations) and notes the omissions in `population$notes`. The
  virtual-cohort weight distribution is anchored to the reference 66 kg with
  log-normal sigma 0.20 to span ~37 - 156 kg.

# Errata

No published erratum or correction was located for PMID 41035281 as of
2026-04-27 (PubMed metadata and the journal landing page carry no
correction-notice link). The items below are *implementation-level*
notational inconsistencies in the source paper that the extraction had to
resolve; they do not affect the final-model parameter point estimates.

- **Abstract vs Table 2 / Eq. 14: CL_DXd units.** The abstract reports
  CL_DXd as `2.66 L/day`; Table 2 and Eq. 14 report the same numerical
  value as `2.66 L/h`. The Table 2 / Eq. 14 unit (L/h) is consistent with
  the small-molecule disposition of DXd (half-life ~5-7 h) and with the
  exatecan congener cited in the Discussion; the abstract value is
  interpreted as a typographical error.
- **Methods text: AST unit label `g/L`.** The Methods reference value for
  AST is given as `22 g/L`. AST is an enzyme activity reported in U/L
  (= IU/L); the canonical AST unit per `inst/references/covariate-columns.md`
  is U/L. The numerical value 22 is consistent with U/L only.
- **DAR equation typesetting.** Equation 1 / Equation 13 uses a piecewise
  inline notation for the `Factor1`-vs-1 cycle scaling that flattens
  ambiguously in plain-text extracts. The MathML in the source XML
  resolves it as `4 * (0.25 + 0.75 * exp(-beta * tad)) * (1 if cycle 1
  else Factor1)` -- the cycle scaling multiplies the entire DAR(t) curve.
