---
title: "Nonracemic amisulpride (Hopkins 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nonracemic amisulpride (Hopkins 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model_load}
ui <- rxode2::rxode(readModelDb("Hopkins_2024_amisulpride"))
```

* Citation: `r ui$reference`
* Article: <https://doi.org/10.1002/cpt.3311>

`r ui$description`

# Population

```{r population}
pop <- ui$population
tibble::tibble(
  Field = names(pop),
  Value = vapply(pop, function(x) paste(
    if (is.null(names(x))) as.character(x) else paste0(names(x), ": ", x),
    collapse = "; "), character(1))
) |>
  knitr::kable(caption = "Population metadata recorded with the model.")
```

The six pooled Phase I studies are listed in Table 1 of the supplement. Table 1
of the paper gives baseline demographics for the four studies with PET imaging.
The analysis dataset comprises 181 subjects, 11,626 plasma samples and 157
D2 receptor occupancy measurements.

# Model structure

The paper's named contribution is the **Distribution Model**: a two-step brain
distribution cascade placed downstream of an ordinary plasma PK model.
Supplement Table 2 of the paper (reproduced as the "Distribution Model" row of
its Table 2) shows that no conventional PD form reproduces all three governing
observations simultaneously; the two-step cascade does.

Absorption is three parallel depots with staggered lag times, representing
regiospecific uptake along the gastrointestinal tract. Each depot's first-order
rate constant is multiplied by a shared Weibull dissolution term whose two
parameters were fitted outside NONMEM to the measured in vitro dissolution curve
of each formulation prototype, so one model spans an oral solution, an
immediate-release tablet, five controlled-release tablet strengths and two
multiparticulate capsules.

```{r model_source, include = FALSE}
src <- readLines(system.file("modeldb/specificDrugs/Hopkins_2024_amisulpride.R",
                            package = "nlmixr2lib"))
closers <- grep("^  \\}\\)$", src)
#' Line range of a top-level `ini({` / `model({` block: its opening line
#' through the first `  })` that follows.
block_span <- function(opener) {
  start <- grep(opener, src)
  stopifnot(length(start) == 1L)
  end <- closers[closers > start][1]
  stopifnot(!is.na(end))
  seq(start, end)
}
stopifnot(length(src) > 100)
```

```{r structure_echo, comment = ""}
cat(paste(src[block_span("^  model\\(\\{$")], collapse = "\n"))
```

# Source trace

Every `ini()` entry in the model file carries an inline comment naming the
table, `$THETA` line or `$OMEGA` line it came from. The table below is parsed
directly out of the installed model file, so it cannot drift from the code.

```{r source_trace}
ini_span <- block_span("^  ini\\(\\{$")
ini_lines <- src[ini_span[-c(1, length(ini_span))]]
keep <- grepl("(<-|~)", ini_lines) & grepl("#", ini_lines)
trace <- tibble::tibble(raw = trimws(ini_lines[keep])) |>
  mutate(
    Parameter = trimws(sub("\\s*(<-|~).*$", "", raw)),
    Definition = trimws(sub("^.*?(<-|~)\\s*", "", sub("\\s*#.*$", "", raw))),
    Source = trimws(sub("^[^#]*#\\s*", "", raw))
  ) |>
  select(Parameter, Definition, Source)
knitr::kable(trace, caption = paste0(
  "Source trace for all ", nrow(trace), " ini() entries, parsed from the model file."))
```

Structural equations trace as follows.

| Model equation | Source |
|---|---|
| `d/dt(depot1..3) <- -kamax_i * wb * depot_i` | Supplement, PK Fitting NONMEM Script, `$DES` DADT(1)-(3) |
| `d/dt(central)` (2-compartment with three absorption inputs) | Supplement, PK script `$DES` DADT(4) |
| `d/dt(peripheral1) <- k12 * central - k21 * peripheral1` | Supplement, PK script `$DES` DADT(5) |
| `wb <- 1 - exp(-(ra * tad(depot1))^gam1)` | Supplement, PK script `$PK`: `WB = 1-EXP((-(KAW*TAD)**GAMAW))` |
| solution bypasses the Weibull term (`wb = 1`) | Supplement, D2 occupancy script `$PK`: `IF (RGN.EQ.4) WB = 1` |
| `tlag3 <- tlag2 + exp(ltlag3 + ...)` | Supplement, PK script `$PK`: `TAL3 = TAL2 + TAL3T` |
| `d/dt(effect1) <- kinf_effect1 * fu * f_esamisulpride * Cc - keff_effect1 * effect1 - kinf_effect2 * effect1` | Supplement, D2 occupancy script `$DES` DADT(6) |
| `d/dt(effect2) <- kinf_effect2 * effect1 - keff_effect2 * effect2` | Supplement, D2 occupancy script `$DES` DADT(7) |
| `D2RO <- 100 * effect2 / (kd + effect2)` | Supplement, D2 occupancy script `$ERROR`: `D2P = (A(7)/(KD + A(7)))*100`; main text Table 2, Distribution Model row |
| `Cc ~ prop(propSd) + add(addSd)` | Supplement, `$ERROR`: `Y = IPRE + IPRE*EPS(1) + EPS(2)` |
| `D2RO ~ add(addSd_D2RO)` | Supplement, D2 occupancy script `$ERROR`: `IF (CMT.EQ.8) Y = IPRE + EPS(3)` |

# Virtual cohort

Three single-dose 200 mg arms are simulated: the oral solution used by the PET
studies, the immediate-release tablet that was carried into Phase II, and the
20% controlled-release tablet, which of the five CR prototypes lies closest to
the peak concentration reported for the CR tablet selected in Study 5.

Each subject receives one dose, so a single occasion is modelled and the
inter-occasion variability terms all draw on `OCC = 1`.

```{r cohort}
form_cols <- c(
  "FORM_SEP4199_CR10", "FORM_SEP4199_CR25", "FORM_SEP4199_IR",
  "FORM_SEP4199_SOLUTION", "FORM_SEP4199_CR15", "FORM_SEP4199_CR25_FED",
  "FORM_SEP4199_MUPS30", "FORM_SEP4199_MUPS225", "FORM_SEP4199_CR20",
  "FORM_SEP4199_CR40"
)

#' Build an event data frame for one formulation arm.
#'
#' Doses go simultaneously into all three parallel absorption compartments, as
#' in the NONMEM dataset; f() and alag() inside the model apply the
#' per-formulation bioavailability fractions and lag times.
make_events <- function(form, dose = 200, n_sub = 1L, times = seq(0, 48, by = 0.1),
                        addl = 0L, ii = 24) {
  stopifnot(form %in% form_cols)
  ev <- rxode2::et(amt = dose, cmt = "depot1", time = 0, addl = addl, ii = ii) |>
    rxode2::et(amt = dose, cmt = "depot2", time = 0, addl = addl, ii = ii) |>
    rxode2::et(amt = dose, cmt = "depot3", time = 0, addl = addl, ii = ii) |>
    rxode2::et(times)
  if (n_sub > 1L) ev <- rxode2::et(ev, id = seq_len(n_sub))
  dat <- as.data.frame(ev)
  # Two endpoints are declared, so observation rows need a dvid; dvid = 1 on
  # every observation returns both Cc and D2RO as columns.
  dat$dvid <- ifelse(dat$evid == 0, 1L, NA_integer_)
  for (nm in form_cols) dat[[nm]] <- 0
  dat[[form]] <- 1
  dat$OCC <- 1
  dat
}

#' Solve, keeping the arm label. useLinCmt = FALSE is required for multi-output
#' models (rxode2's automatic ODE-to-linCmt conversion corrupts the dvid map).
solve_arm <- function(mod, form, label, ...) {
  out <- rxode2::rxSolve(mod, make_events(form, ...), returnType = "data.frame",
                         useLinCmt = FALSE)
  if (is.null(out$id)) out$id <- 1L
  out$treatment <- label
  out
}

arms <- c(Solution = "FORM_SEP4199_SOLUTION",
          `IR tablet` = "FORM_SEP4199_IR",
          `CR tablet (20%)` = "FORM_SEP4199_CR20")

n_per_arm <- 60L
# Re-seed before EACH arm so all three draw the identical sequence of etas.
# That makes subject i the same virtual person in every arm, which turns the
# across-arm contrasts below into paired comparisons. Seeding once outside the
# loop would give each arm a different cohort and silently unpair them.
cohort <- lapply(names(arms), function(lbl) {
  rxode2::rxSetSeed(20240501)
  solve_arm(ui, arms[[lbl]], lbl, n_sub = n_per_arm)
}) |>
  bind_rows() |>
  mutate(treatment = factor(treatment, levels = names(arms)))

# Confirm the arms really are paired: with common random numbers the individual
# clearance draws must be identical subject-by-subject across arms.
cl_by_arm <- cohort |>
  distinct(treatment, id, cl) |>
  tidyr::pivot_wider(names_from = treatment, values_from = cl)
stopifnot(
  nrow(cl_by_arm) == n_per_arm,
  isTRUE(all.equal(cl_by_arm[[2]], cl_by_arm[[3]])),
  isTRUE(all.equal(cl_by_arm[[2]], cl_by_arm[[4]]))
)

obs <- cohort |> filter(time > 0)
stopifnot(nrow(obs) > 0, !anyNA(obs$Cc), !anyNA(obs$D2RO), all(obs$Cc >= 0))
```

Typical-value profiles use `rxode2::zeroRe()` so the curves are the
population prediction rather than one realisation.

```{r typical}
tv <- rxode2::zeroRe(ui)
```

# Replicating the published observations

## Hysteresis (Figure 1a)

Plasma concentrations wash out within about 24 h while D2 receptor occupancy
needs about 5 days. Both series are scaled to their own peak, as in the source
figure.

```{r fig_hysteresis}
hyst <- solve_arm(tv, "FORM_SEP4199_IR", "IR tablet", times = seq(0, 144, by = 0.25)) |>
  filter(time > 0) |>
  transmute(time,
            `Plasma concentration` = Cc / max(Cc),
            `D2 receptor occupancy` = D2RO / max(D2RO)) |>
  pivot_longer(-time, names_to = "series", values_to = "fraction")

ggplot(hyst, aes(time, fraction, colour = series)) +
  geom_line(linewidth = 0.9) +
  scale_x_continuous(breaks = seq(0, 144, by = 24)) +
  labs(x = "Time (h)", y = "Fraction of own peak", colour = NULL,
       title = "Replicates Figure 1a of Hopkins 2024",
       subtitle = "200 mg IR tablet, typical-value prediction") +
  theme_bw() + theme(legend.position = "bottom")
```

```{r gate_hysteresis}
h <- solve_arm(tv, "FORM_SEP4199_IR", "IR tablet", times = seq(0, 144, by = 0.25)) |>
  filter(time > 0)
at <- function(v, tt) v[h$time == tt]
tmax_pk <- h$time[which.max(h$Cc)]
tmax_ro <- h$time[which.max(h$D2RO)]
pk_frac <- h$Cc / max(h$Cc)
ro_frac <- h$D2RO / max(h$D2RO)
c(tmax_pk = tmax_pk, tmax_ro = tmax_ro,
  pk_frac_at_24h = at(pk_frac, 24),
  ro_frac_at_8h = at(ro_frac, 8), ro_frac_at_24h = at(ro_frac, 24),
  ro_frac_at_27h = at(ro_frac, 27), ro_frac_at_120h = at(ro_frac, 120))

stopifnot(
  # Occupancy peaks well after the plasma peak -- the definition of hysteresis.
  tmax_ro > 3 * tmax_pk,
  # "the plasma PK (24-hour washout)": plasma is essentially gone at 24 h.
  at(pk_frac, 24) < 0.10,
  # "maximal D2R receptor occupancies were achieved by 8 hours and maintained
  # out to 27 hours postdose" -- occupancy is within 20% of its own peak across
  # that whole window while plasma falls by more than an order of magnitude.
  at(ro_frac, 8) > 0.80, at(ro_frac, 24) > 0.80, at(ro_frac, 27) > 0.75,
  # "up to 5 days were required for complete washout of D2R occupancy levels."
  at(ro_frac, 120) < 0.05
)
```

## Lack of accumulation over 7 daily doses (Figure 1c)

```{r fig_accumulation}
md <- lapply(c("IR tablet", "CR tablet (20%)"), function(lbl) {
  solve_arm(tv, arms[[lbl]], lbl, times = seq(0, 192, by = 0.25), addl = 6L, ii = 24)
}) |>
  bind_rows() |>
  filter(time > 0)

pet <- md |> filter(time %in% c(27, 27 + 144))

ggplot(md, aes(time, D2RO, colour = treatment)) +
  geom_line(linewidth = 0.8) +
  geom_point(data = pet, size = 2.6) +
  scale_x_continuous(breaks = seq(0, 192, by = 24)) +
  labs(x = "Time (h)", y = "D2 receptor occupancy (%)", colour = NULL,
       title = "Replicates Figure 1c of Hopkins 2024",
       subtitle = paste("200 mg once daily for 7 days; points are the PET scan times",
                        "(27 h after dose 1 and after dose 7)")) +
  theme_bw() + theme(legend.position = "bottom")
```

```{r gate_accumulation}
acc <- pet |>
  mutate(scan = ifelse(time == 27, "day1", "day7")) |>
  select(treatment, scan, D2RO) |>
  pivot_wider(names_from = scan, values_from = D2RO) |>
  mutate(ratio = day7 / day1)
knitr::kable(acc, digits = 2,
             caption = "Occupancy 27 h after the first and the seventh daily dose.")

stopifnot(
  # The paper's claim is that occupancy does not accumulate SUBSTANTIALLY. The
  # initial indirect-response model it replaced predicted the day-7 scan roughly
  # double the day-1 scan (gray curves of Figure 1c); the Distribution Model
  # keeps the rise modest.
  all(acc$ratio > 1), all(acc$ratio < 1.5),
  # Both formulations reach a similar steady-state occupancy: the paper's
  # "pharmacodynamic bioequivalence in the brain" claim.
  abs(diff(acc$day7)) / mean(acc$day7) < 0.25
)
```

## Dose-response without receptor saturation (Figure 1b)

```{r fig_doseresponse}
dr <- lapply(c(200, 300, 400, 600, 700), function(d) {
  s <- solve_arm(tv, "FORM_SEP4199_IR", "IR tablet", dose = d, times = seq(0, 48, by = 0.25))
  tibble::tibble(dose = d, rod2_27h = s$D2RO[s$time == 27])
}) |>
  bind_rows()

ggplot(dr, aes(dose, rod2_27h)) +
  geom_line(linewidth = 0.8) + geom_point(size = 2.5) +
  ylim(0, 100) +
  labs(x = "SEP-4199 dose (mg)", y = "D2 receptor occupancy at 27 h (%)",
       title = "Replicates the dose-response of Figure 1b of Hopkins 2024",
       subtitle = "IR tablet, typical-value prediction; Study 3 spanned 200-700 mg") +
  theme_bw()

knitr::kable(dr, digits = 1, col.names = c("Dose (mg)", "Occupancy at 27 h (%)"))
```

```{r gate_doseresponse}
stopifnot(
  # Monotone in dose.
  all(diff(dr$rod2_27h) > 0),
  # Not saturated: the paper states receptor is not saturated at the tested dose
  # levels, and targets occupancy below 50% at the 200/400 mg therapeutic doses.
  max(dr$rod2_27h) < 70,
  dr$rod2_27h[dr$dose == 400] < 50,
  # A 3.5-fold dose increase still produces a substantial occupancy increase,
  # i.e. the curve has not flattened.
  dr$rod2_27h[dr$dose == 700] / dr$rod2_27h[dr$dose == 200] > 1.8
)
```

## Controlled release lowers peak plasma concentration (Figures 4b and 5)

```{r fig_ir_vs_cr}
med <- obs |>
  group_by(treatment, time) |>
  summarise(median = median(Cc),
            lo = quantile(Cc, 0.1), hi = quantile(Cc, 0.9), .groups = "drop")

ggplot(med, aes(time, median, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.9) +
  coord_cartesian(xlim = c(0, 24)) +
  labs(x = "Time (h)", y = "Total amisulpride (ng/mL)", colour = NULL, fill = NULL,
       title = "Replicates Figure 4b / Figure 5 (PK panel) of Hopkins 2024",
       subtitle = paste0("200 mg single dose, ", n_per_arm,
                         " subjects per arm; median with 10th-90th percentile band")) +
  theme_bw() + theme(legend.position = "bottom")
```

# Noncompartmental analysis

```{r pknca}
# Built from `cohort`, NOT from `obs`: `obs` drops time == 0, and a PKNCA
# concentration frame with no time-zero record triggers "Requesting an AUC range
# starting (0) before the first measurement" once per subject.
conc_data <- cohort |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)
stopifnot(all(tapply(conc_data$time, conc_data$treatment, min) == 0))

dose_data <- conc_data |>
  distinct(id, treatment) |>
  mutate(time = 0, dose = 200)

o_conc <- PKNCA::PKNCAconc(conc_data, Cc ~ time | treatment + id,
                           concu = "ng/mL", timeu = "h")
o_dose <- PKNCA::PKNCAdose(dose_data, dose ~ time | treatment + id, doseu = "mg")
o_data <- PKNCA::PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(start = 0, end = 24,
                         cmax = TRUE, tmax = TRUE, auclast = TRUE, half.life = TRUE)
)
nca_raw <- suppressWarnings(PKNCA::pk.nca(o_data))

nca_res <- as.data.frame(nca_raw) |>
  filter(!is.na(PPORRES)) |>
  group_by(treatment, PPTESTCD) |>
  summarise(value = ifelse(unique(PPTESTCD) == "tmax",
                           median(PPORRES), exp(mean(log(PPORRES)))),
            .groups = "drop") |>
  pivot_wider(names_from = PPTESTCD, values_from = value)

stopifnot(nrow(nca_res) == length(arms), all(c("cmax", "tmax", "auclast") %in% names(nca_res)))
knitr::kable(nca_res, digits = c(0, 1, 2, 0, 2),
             caption = "Simulated NCA: geometric mean (median for tmax) per arm.")
```

## Comparison against the published values

The paper reports three summary statistics that this model can be scored
against: the median AUC0-24 of the 200 mg liquid arms of Studies 3 and 4
(Figure 3 caption), and the geometric mean peak concentrations of the
immediate-release and controlled-release tablets in Study 5 (Figure 5). Tmax and
half-life are not reported numerically, so they are omitted from the comparison
rather than compared against a value read off a figure.

```{r nca_compare}
published <- tibble::tribble(
  ~treatment,        ~cmax, ~auclast,
  "Solution",        NA,    3000,
  "IR tablet",       599,   NA,
  "CR tablet (20%)", 314,   NA
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "treatment",
  units         = c(cmax = "ng/mL", auclast = "ng*h/mL"),
  tolerance_pct = 20
)

# ncaComparisonTable() already names the header column "NCA parameter" and
# renders an unmatched cell as a dash, so no renaming is needed here.
knitr::kable(cmp,
             caption = paste("Simulated vs published.",
                             "* marks a difference above 20%."),
             align = c("l", "l", "r", "r", "r"))
```

```{r gate_nca}
gm <- function(tr, pp) nca_res[[pp]][nca_res$treatment == tr]
sol_auc <- gm("Solution", "auclast")
ir_cmax <- gm("IR tablet", "cmax")
cr_cmax <- gm("CR tablet (20%)", "cmax")
c(solution_auc0_24 = sol_auc, ir_cmax = ir_cmax, cr_cmax = cr_cmax,
  cr_over_ir = cr_cmax / ir_cmax)

stopifnot(
  # Exposure is the parameter the model reproduces well: the 200 mg liquid arm
  # lands within 20% of the published median AUC0-24 of 3000 ng*h/mL.
  abs(sol_auc / 3000 - 1) < 0.20,
  # The CR peak concentration matches the published geometric mean closely.
  abs(cr_cmax / 314 - 1) < 0.20,
  # The IR peak is under-predicted -- a limitation the supplement itself records
  # ("slight under-prediction of the highest concentrations was somewhat
  # apparent"). This gate pins the size of that gap so it cannot silently drift.
  ir_cmax / 599 > 0.55, ir_cmax / 599 < 0.80,
  # Controlled release still lowers the peak relative to immediate release,
  # which is the finding the formulation programme rested on.
  cr_cmax < ir_cmax
)
```

Per-subject paired checks are more robust than cohort extremes, because the
absorption terms carry large inter-individual and inter-occasion variability and
the extreme of a random cohort is not reproducible across rxode2 versions.

```{r gate_paired}
paired <- obs |>
  group_by(treatment, id) |>
  summarise(cmax = max(Cc), tmax_pk = time[which.max(Cc)],
            tmax_ro = time[which.max(D2RO)], .groups = "drop")

# Hysteresis holds subject by subject, not just on average.
stopifnot(all(paired$tmax_ro - paired$tmax_pk > 2))

cr_vs_ir <- paired |>
  filter(treatment != "Solution") |>
  select(treatment, id, cmax) |>
  pivot_wider(names_from = treatment, values_from = cmax) |>
  mutate(ratio = `CR tablet (20%)` / `IR tablet`)

quantile(cr_vs_ir$ratio, c(0.1, 0.5, 0.9))

stopifnot(
  # Centre, not extreme: the controlled-release tablet lowers the peak
  # concentration in the typical subject.
  median(cr_vs_ir$ratio) < 0.95,
  # Robust envelope rather than max(): the extreme of a random cohort is not
  # reproducible across rxode2 versions, so the 90th percentile is asserted
  # instead. Even the least-affected decile is not materially above parity.
  quantile(cr_vs_ir$ratio, 0.9) < 1.10
)
```

The direction of the controlled-release effect is reproduced, but its magnitude
is not: the simulated median CR-to-IR peak ratio is about 0.89 against the
0.52 implied by the published geometric means of 314 and 599 ng/mL. The gap is
entirely on the immediate-release side, whose peak this model under-predicts by
about a third for the reason recorded in the first Errata bullet below; the CR
peak itself lands within a few percent of the published value. Nothing was
adjusted to narrow it.

# Assumptions and deviations

* **Which published parameter set the model encodes.** The supplement reports
  the PK layer twice and the two reports disagree. Its Table 2 ("NONMEM PK
  Fitting Results") gives CL 128 L/h, Vc 203 L, ALAG2 0.97 h and a distinct
  IIV / IOV structure; the PK control stream printed in the same supplement
  carries `$THETA ... FIX` values of CL 106 L/h, Vc 289 L, ALAG2 0.241 h with a
  different IIV / IOV structure. This model encodes the **control stream**, for
  four reasons. (i) Table 2's IIV column is not usable: it reports variances up
  to 39.1 on parameters with exponential inter-individual variability, which
  corresponds to a coefficient of variation of order 1e8 and cannot be
  simulated; the supplement itself notes that "obtaining full convergence was
  typically not feasible". The control stream's variances (0.105 on CL, i.e.
  33% CV) are ordinary. (ii) The control stream is the executable specification
  whose `$TABLE` writes exactly the per-subject columns that the D2 receptor
  occupancy script reads in `$INPUT`, so it is the PK that underlies the
  published Distribution Model results. (iii) It is internally self-consistent
  across `$THETA`, `$OMEGA` and the IOV block. (iv) Scored against the paper's
  own published anchors it fits better than Table 2: median AUC0-24 for the
  200 mg liquid arm 3395 vs 3588 ng*h/mL against a published 3000, and IR
  typical-value Cmax 381 vs 323 ng/mL against a published 599. Nothing was
  tuned; both sets were simulated as printed and compared.
* **`ALAG3` is an increment, not an absolute lag.** The control stream computes
  `TAL3 = TAL2 + TAL3T`, so `THETA(6)` is the additional delay of the third
  absorption compartment beyond the second. The absolute lag is
  0.241 + 2.4 = 2.641 h. Reading `THETA(6)` as an absolute lag would place the
  third absorption wave an hour and a half too early.
* **Unit conversion on the plasma concentration.** `$ERROR` computes
  `CT = A(4)/V4` with no scaling, which means the NONMEM dataset carried dose
  amounts in ug. Doses are declared here in mg so that the published dose levels
  can be entered directly, so `model()` applies an explicit factor of 1000 to
  convert mg/L to ng/mL. It is a unit conversion, not a fitted parameter, and it
  applies before the effect compartments because their influx is driven by the
  plasma concentration on the same ng/mL scale as `kd = 9 ng/mL`.
* **PD parameters come from Table 3, not from the occupancy control stream.**
  That script's `$THETA` entries (KE1 0.0723, KE2 0.221, KE3 0.0605, KT 0.0741)
  are bounded initial estimates for the run; the final values are supplement
  Table 3 (KE1 0.0221, KE2 0.0685, KE3 0.0602, KT 0.42). The fixed KD, FU and SP
  agree between the two.
* **The Weibull dissolution constants come from the executed `$PK` block.**
  Supplement Table 5 rounds two of them differently from the code that was run
  (RGN9 `TKA1` printed as 0.80 against 0.8457 in `$PK`, RGN10 `TGAMA1` printed
  as 0.70 against 0.7764). The four-decimal values in `$PK` are used. The
  supplement also states that the pairs for the fed 25% CR tablet and for the
  solution were interpolated rather than fitted to a dissolution curve.
* **The oral solution bypasses the Weibull term.** The PK script applies a
  nominal fast Weibull pair to the solution while the D2 occupancy script sets
  `WB = 1` for it. This model follows the occupancy script, which is the final
  integrated model and is also the physically sensible reading (a solution has
  no solid-state dissolution step). The numerical difference is confined to the
  first half hour after a solution dose.
* **Zero-variance IIV terms are omitted.** The control stream fixes the
  inter-individual variances of `KA1`-`KA3` and `F1`-`F3` to exactly zero. No
  eta is carried for them, because a zero-variance eta makes OMEGA singular and
  `chol()` fails during sampling. The inter-occasion variability on `F1`-`F3`
  is non-zero and is retained.
* **Which CR prototype Study 5 selected is not stated.** The paper says a single
  CR prototype was selected "based on PK parameters evaluated in previous
  cohorts" but does not name it. The 20% CR tablet is used here because its
  simulated peak concentration is the closest of the five CR prototypes to the
  published 314 ng/mL. All ten formulations are available in the model through
  their `FORM_SEP4199_*` indicators.
* **The IR peak concentration is under-predicted by about a third.** This is a
  property of the published model, not of the transcription: the supplement
  records that "slight under-prediction of the highest concentrations was
  somewhat apparent", and the main text emphasises the "short-lasting peak
  concentrations with the IR formulation" that a three-depot Weibull absorption
  model necessarily smooths. Exposure (AUC) and the CR peak are reproduced
  within 20%. No parameter was adjusted to close the gap.
* **Sex and food are carried as `covariatesDataExcluded`.** `SEX` and `FED`
  appear in the NONMEM `$INPUT` records but are never referenced in `$PK` or
  `$DES`. The single fed arm is carried structurally instead, as its own
  regimen indicator with its own bioavailability triple and Weibull pair.
* **`sex_female_pct` is deliberately absent from `population`.** Table 1 reports
  sex per study rather than pooled; the ascending-single-dose study enrolled male
  subjects only per its title; and no sex breakdown is published for Study 2 or
  for Study 5 as a whole. A pooled female percentage across the 181 subjects
  therefore cannot be computed. Women were enrolled -- Study 4 was 36% female and
  Study 5 Parts 1 and 2 were 47% and 56% female -- so recording 0 would assert an
  all-male cohort that the source contradicts.
* **Models the paper rejected are not extracted.** Table 2 of the paper lists
  five alternative PD forms (linear direct effect, saturated Emax, effect
  compartment, concentration-difference, ratio) with their equations but no
  parameter estimates, and the earlier indirect-response model built on
  single-dose data is described without published parameters. None of them can
  be encoded without inventing values, so only the Distribution Model is
  packaged. The concentration-QT relationship of Figure 4c and the
  `Cmax`-occupancy regression of Figure 3b are likewise reported graphically
  without parameter values.
