---
title: "Furosemide (Hornik 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Furosemide (Hornik 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Hornik_2025_furosemide"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1007/s40262-025-01515-2>
* Supplement (Table S1, Figures S1-S8): <https://doi.org/10.1007/s40262-025-01515-2> (electronic supplementary material)

Hornik and colleagues developed an adult population PK model for furosemide from
a randomized crossover study comparing a buffered subcutaneous formulation
(Furoscix, delivered by a wearable On-Body Infusor) against intravenous
furosemide, then scaled that model allometrically to predict exposure in
adolescents receiving the fixed 80 mg subcutaneous regimen.

# Population

The adult model was built on 15 adults with chronic heart failure (NYHA class
II/III) and chronic volume overload enrolled in NCT02329834 (Sect. 2.1,
Sect. 3.1). Sixteen adults received furosemide; one was excluded for high
pre-dose furosemide concentrations. Median age was 69 years (range 52-83) and 14
of 15 (93.3%) were men. Each patient received both treatments in a crossover
with a 7-day washout: 80 mg subcutaneously (30 mg over the first 60 min, then
12.5 mg/h for 4 h) and 80 mg intravenously (40 mg over 2 min, repeated 2 h
later). Median sampling was 21 samples per patient on the intravenous occasion
and 18 on the subcutaneous occasion.

The paper reports no baseline demographics table, so weight, body mass index and
renal function are available only from the axes of the ETA-versus-covariate
panels in the supplement (Figures S1-S5): weight roughly 60-120 kg, creatinine
clearance roughly 60-140 mL/min/1.73 m^2, body mass index roughly 25-35 kg/m^2
and serum creatinine roughly 1.0-1.8 mg/dL. The median renal function used as
the covariate reference is 86 (Sect. 2.6).

```{r population}
str(readModelDb("Hornik_2025_furosemide")()$population)
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Hornik_2025_furosemide.R` carries an in-file comment
naming its origin. They are collected here for review. Equation numbers are
those of the published article.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` (Ka)   | 1.30 1/h | Table 1, "K a , h-1"; structure Eq. 14 |
| `lcl` (CL)   | 6.51 L/h | Table 1, "CL, L/h"; structure Eq. 15 |
| `lvc` (Vc)   | 5.37 L   | Table 1, "V c , L"; structure Eq. 16 |
| `lq` (Q)     | 3.74 L/h | Table 1, "Q , L/h"; structure Eq. 17 |
| `lvp` (Vp)   | 5.95 L   | Table 1, "V p , L"; structure Eq. 18 |
| `lfdepot` (F, subcutaneous) | 0.96 | Sect. 3.2 text; logit form Eq. 19. Not in Table 1 -- see "Reconciling the reported bioavailability" below |
| `e_crcl_cl`  | 0.74 | Table 1, "eGFR on CL"; power form Eq. 15, reference 86 |
| `e_wt_cl_q`  | 0.75 (fixed) | Eqs. 10 and 12; held at the conventional value per Sect. 2.6 and Sect. 4 |
| `e_wt_vc_vp` | 1 (fixed)    | Eqs. 11 and 13 |
| `etalka`  | 21.7% CV | Table 1, IIV block (shrinkage 10%) |
| `etalcl`  | 18.5% CV | Table 1, IIV block (shrinkage 0%) |
| `etalvc`  | 31.7% CV | Table 1, IIV block (shrinkage 0%) |
| `etalq`   | 33.1% CV | Table 1, IIV block (shrinkage 6%) |
| `etalvp`  | 18.9% CV | Table 1, IIV block (shrinkage 3%) |
| `propSd`  | 0.126 | Table 1, "Proportional error, %" = 12.6%; combined form Eq. 5 |
| `addSd`   | 0.100 ug/mL | Table 1, "Additive error, ng/mL" = 100 ng/mL |
| Exponential IIV, `exp(eta)` | n/a | Eq. 1 |
| Combined residual error | n/a | Eq. 5 |
| Two-compartment ODEs with first-order subcutaneous absorption | n/a | Sect. 3.2 ("two-compartment model with linear kinetics and a first-order absorption rate constant") |
| Allometric scaling to 70 kg | n/a | Eqs. 9-13 |

Covariate effects that were screened but **not** retained (age on Vp, serum
creatinine, body mass index, sex, NYHA class) are recorded in the model file's
`covariatesDataExcluded` metadata rather than implemented; the covariate search
is Table S1 and Figures S1-S5 of the supplement.

The IIV variances are stored on the variance scale as
`omega^2 = log(CV^2 + 1)`:

```{r iiv-conversion}
tibble::tibble(
  parameter = c("etalka", "etalcl", "etalvc", "etalq", "etalvp"),
  cv_pct    = c(21.7, 18.5, 31.7, 33.1, 18.9)
) |>
  dplyr::mutate(omega_sq = round(log((cv_pct / 100)^2 + 1), 6)) |>
  dplyr::rename("Parameter" = parameter, "Table 1 CV%" = cv_pct,
                "omega^2 in ini()" = omega_sq) |>
  knitr::kable(caption = "Table 1 CV% converted to the variance scale.")
```

# Virtual cohort

The original observed concentrations are not public. The simulations below
rebuild the paper's virtual adolescent population (Sect. 2.7): three weight
bands (42.5-50.0, >50-60 and >60-70 kg) with weight distributed uniformly within
each band, plus a typical 70 kg adult reference. Renal function is fixed at the
adult population median of 86 for every virtual subject, which is what the
authors did -- Sect. 4 states the pediatric parameters were fixed at "the
population parameter estimate of a 70-kg adult with an eGFR of 86 mL/min".
Between-subject variability is switched off, again matching Sect. 2.7 ("To
represent the anticipated exposures in typical adolescents within a given weight
group, IIV was not included in the simulations"), so the only source of spread
within a band is body weight.

The paper simulated 1000 virtual adolescents per band; 200 per arm is used here
(the library cap), which is more than sufficient because the weight grid is
deterministic.

```{r cohort}
set.seed(20250513)

n_per_band <- 200L

bands <- tibble::tribble(
  ~grp,             ~wt_lo, ~wt_hi,
  "42.5-50.0 kg",    42.5,   50.0,
  ">50-60 kg",       50.0,   60.0,
  ">60-70 kg",       60.0,   70.0,
  "Adult (70 kg)",   70.0,   70.0
)

# Observations every 0.1 h to 24 h (the paper's grid, Sect. 2.7). The grid is
# extended to 72 h at a coarser spacing so the AUC mass-balance identity below
# can be checked over a practically infinite window; the 0-24 h interval used
# for the comparison against Table 2 is unaffected.
obs_times <- c(seq(0, 24, by = 0.1), seq(24.5, 72, by = 0.5))

make_band <- function(grp, wt_lo, wt_hi, id_offset) {
  wt <- if (isTRUE(all.equal(wt_lo, wt_hi))) {
    rep(wt_lo, n_per_band)
  } else {
    seq(wt_lo, wt_hi, length.out = n_per_band)
  }
  subj <- tibble::tibble(
    id   = id_offset + seq_len(n_per_band),
    WT   = wt,
    CRCL = 86,
    grp  = grp
  )
  # 80 mg subcutaneously: 30 mg over the first hour, then 12.5 mg/h for 4 h.
  # `dur` (not `rate`) is used so bioavailability scales the delivered rate
  # while the pump's programmed duration stays fixed.
  doses <- dplyr::bind_rows(
    dplyr::mutate(subj, time = 0, amt = 30, evid = 1L, dur = 1, cmt = "depot"),
    dplyr::mutate(subj, time = 1, amt = 50, evid = 1L, dur = 4, cmt = "depot")
  )
  obs <- tidyr::crossing(subj, time = obs_times) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, dur = NA_real_, cmt = "central")
  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  lapply(seq_len(nrow(bands)), function(i) {
    make_band(bands$grp[i], bands$wt_lo[i], bands$wt_hi[i],
              id_offset = (i - 1L) * 1000L)
  })
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
stopifnot(nrow(dplyr::distinct(events, id)) == 4L * n_per_band)
```

Each band's dosing totals the fixed 80 mg regimen:

```{r dose-check}
events |>
  dplyr::filter(evid == 1) |>
  dplyr::group_by(grp, id) |>
  dplyr::summarise(total_dose_mg = sum(amt), .groups = "drop") |>
  dplyr::group_by(grp) |>
  dplyr::summarise(total_dose_mg = mean(total_dose_mg), .groups = "drop") |>
  dplyr::rename("Weight band" = grp, "Total dose (mg)" = total_dose_mg) |>
  knitr::kable(caption = "Fixed 80 mg subcutaneous dose in every band.")
```

# Simulation

```{r simulate}
mod <- readModelDb("Hornik_2025_furosemide")

# Sect. 2.7: IIV was excluded from the dosing simulations.
mod_typical <- rxode2::zeroRe(mod)

sim <- rxode2::rxSolve(
  mod_typical, events = events,
  keep = c("WT", "grp"),
  returnType = "data.frame"
) |>
  as.data.frame()

stopifnot(dplyr::n_distinct(sim$id) == 4L * n_per_band)
```

With the random effects zeroed, two subjects at the same weight must give
identical predictions -- the adult band is 200 copies of a 70 kg subject, so its
spread must be exactly zero:

```{r determinism-check}
adult_cmax <- sim |>
  dplyr::filter(grp == "Adult (70 kg)", !is.na(Cc)) |>
  dplyr::group_by(id) |>
  dplyr::summarise(cmax = max(Cc), .groups = "drop")

stopifnot(nrow(adult_cmax) == n_per_band)
stopifnot(diff(range(adult_cmax$cmax)) == 0)
cat("Adult band spread in Cmax:", diff(range(adult_cmax$cmax)), "ug/mL\n")
```

# Replicate published figures

```{r figure-2, fig.cap = "Replicates Figure 2 of Hornik 2025."}
sim |>
  dplyr::filter(!is.na(Cc), time <= 24) |>
  dplyr::group_by(grp, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50 * 1000, colour = grp, fill = grp)) +
  geom_ribbon(aes(ymin = Q05 * 1000, ymax = Q95 * 1000), alpha = 0.2,
              colour = NA) +
  geom_line(linewidth = 0.7) +
  labs(
    x = "Time since start of subcutaneous administration (h)",
    y = "Furosemide plasma concentration (ug/L)",
    colour = NULL, fill = NULL,
    title = "Figure 2 - simulated adolescent profiles versus a typical 70 kg adult",
    caption = "Replicates Figure 2 of Hornik 2025. Lines are medians; bands are the 5th-95th percentiles across body weight."
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

The shape matches the published figure: concentrations rise steeply during the
30 mg loading hour, peak shortly after the infusion steps down to 12.5 mg/h at
1 h, decline slowly through the remainder of the 5 h infusion, and then fall
away once administration stops. Exposure increases as body weight decreases,
because the dose is fixed while clearance scales with weight.

# Structural identity check

Because clearance is linear and the dose is a finite infusion, total exposure
must satisfy `AUC(0-inf) = F * Dose / CL` exactly for every subject. Checking
this per subject, rather than comparing group medians, is a much stricter test:
it exercises the allometric scaling, the bioavailability term and the two-
compartment disposition together.

```{r identity-check}
per_subject <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(id, grp, WT, cl) |>
  dplyr::summarise(
    auc_0_72 = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    .groups  = "drop"
  ) |>
  dplyr::mutate(
    fdepot   = 0.96,
    expected = fdepot * 80 / cl,
    ratio    = auc_0_72 / expected
  )

summary(per_subject$ratio)

# AUC(0-72 h) is a >30 half-life window, so the truncation loss is negligible.
stopifnot(all(abs(per_subject$ratio - 1) < 0.001))
```

Clearance itself must follow Eq. 10 exactly, `CL = 6.51 * (WT/70)^0.75`, since
renal function is held at the reference value:

```{r cl-identity}
stopifnot(all(abs(per_subject$cl / (6.51 * (per_subject$WT / 70)^0.75) - 1) < 1e-8))

# Sect. 3.3 reports weight-normalized clearance of 1.55 mL/min/kg.
cl_norm_adult <- with(
  dplyr::filter(per_subject, grp == "Adult (70 kg)")[1, ],
  cl * 1000 / 60 / WT
)
cat(sprintf("Weight-normalized CL at 70 kg: %.3f mL/min/kg (paper: 1.55)\n",
            cl_norm_adult))
stopifnot(abs(cl_norm_adult - 1.55) < 0.005)
```

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, grp)

# Guarantee a time = 0 record per subject; pre-dose concentration is zero for
# this extravascular regimen.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, grp) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, grp, time, .keep_all = TRUE) |>
  dplyr::arrange(id, grp, time)

conc_obj <- PKNCA::PKNCAconc(
  sim_nca, Cc ~ time | grp + id,
  concu = "ug/mL", timeu = "h"
)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, grp)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | grp + id, doseu = "mg")

# Sect. 2.7 / Table 2 report AUC(0-24) and Cmax over the 24 h window.
intervals <- data.frame(
  start     = 0,
  end       = 24,
  cmax      = TRUE,
  tmax      = TRUE,
  auclast   = TRUE,
  half.life = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)
```

```{r pknca-summary}
nca_res$result |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "half.life")) |>
  dplyr::group_by(grp, PPTESTCD) |>
  dplyr::summarise(
    geomean = exp(mean(log(PPORRES))),
    cv_pct  = sqrt(exp(stats::var(log(PPORRES))) - 1) * 100,
    .groups = "drop"
  ) |>
  dplyr::mutate(
    PPTESTCD = nlmixr2lib::ncaParamLabel(PPTESTCD),
    geomean  = signif(geomean, 4),
    cv_pct   = round(cv_pct, 2)
  ) |>
  dplyr::rename("Weight band" = grp, "NCA parameter" = PPTESTCD,
                "Geometric mean" = geomean, "CV%" = cv_pct) |>
  knitr::kable(caption = "Simulated NCA over 0-24 h (Cc in ug/mL, time in h).")
```

## Comparison against published NCA

Table 2 of Hornik 2025 reports AUC(0-24) in ug*h/L and Cmax in ug/L. The model
works in ug/mL (dose in mg, volumes in L), so the published values are divided
by 1000 below: 16,800 ug*h/L is 16.8 ug*h/mL, 2590 ug/L is 2.59 ug/mL, and so
on.

```{r nca_compare}
published <- tibble::tribble(
  ~grp,             ~auclast, ~cmax,
  "42.5-50.0 kg",   16.8,     2.590,
  ">50-60 kg",      14.7,     2.240,
  ">60-70 kg",      13.0,     1.960,
  "Adult (70 kg)",  12.4,     1.900
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "grp",
  units         = c(cmax = "ug/mL", auclast = "ug*h/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated versus Table 2 of Hornik 2025. * differs from reference by >20%.",
  align   = c("l", "l", "r", "r", "r")
)
```

```{r nca_compare_footnote, echo = FALSE, results = "asis"}
fn <- attr(cmp, "footnote")
if (!is.null(fn)) cat(fn, "\n")
```

No row exceeds the 20% tolerance. Every simulated value nonetheless sits about
4% below the published one, and the offset is almost identical for AUC and for
Cmax across all four groups -- the signature of a single multiplicative factor
rather than a structural disagreement. That factor is bioavailability, as the
next section shows.

# Reconciling the reported bioavailability

Section 3.2 states that "bioavailability following subcutaneous furosemide
administration was estimated at 0.96", and that is the value packaged in the
model. Table 1 does not list the corresponding `theta(6)`, so 0.96 is the only
published number for this parameter.

The paper's own simulations, however, are only reproducible with `F = 1`. Two
independent internal checks establish this.

First, clearance is pinned by Sect. 3.3: the reported weight-normalized
clearance of 1.55 mL/min/kg is exactly `6.51 L/h / 70 kg`, confirming that the
simulations used the Table 1 estimate of 6.51 L/h and the 70 kg reference.

Second, with clearance fixed, `AUC(0-24) ~ F * Dose / CL` back-solves
bioavailability directly from each published exposure. Since the truncation loss
over 24 h is negligible (below), the arithmetic is essentially exact:

```{r f-backsolve}
band_gm_wt <- function(a, b) {
  if (isTRUE(all.equal(a, b))) return(a)
  exp((b * log(b) - b - a * log(a) + a) / (b - a))  # geometric mean of U(a, b)
}

backsolve_f <- bands |>
  dplyr::mutate(
    gm_wt     = mapply(band_gm_wt, wt_lo, wt_hi),
    cl        = 6.51 * (gm_wt / 70)^0.75,
    published = c(16.8, 14.7, 13.0, 12.4),          # ug*h/mL, Table 2
    f_implied = published * cl / 80
  )

backsolve_f |>
  dplyr::mutate(dplyr::across(c(gm_wt, cl, f_implied), \(x) round(x, 3))) |>
  dplyr::select(grp, gm_wt, cl, published, f_implied) |>
  dplyr::rename(
    "Weight band"                = grp,
    "Geometric mean weight (kg)" = gm_wt,
    "CL (L/h)"                   = cl,
    "Published AUC0-24 (ug*h/mL)" = published,
    "Implied F"                  = f_implied
  ) |>
  knitr::kable(caption = "Bioavailability back-solved from Table 2.")
```

The three adolescent bands -- where Sect. 2.7 guarantees that weight is the only
source of variation -- all return `F` within a fraction of a percent of 1.00,
not 0.96. Rescaling the simulated exposures by `1 / 0.96` therefore reproduces
Table 2 almost exactly:

```{r f-rescale}
rescaled <- nca_res$result |>
  dplyr::filter(PPTESTCD %in% c("auclast", "cmax")) |>
  dplyr::group_by(grp, PPTESTCD) |>
  dplyr::summarise(
    geomean = exp(mean(log(PPORRES))),
    cv_pct  = sqrt(exp(stats::var(log(PPORRES))) - 1) * 100,
    .groups = "drop"
  ) |>
  dplyr::mutate(at_f1 = geomean / 0.96) |>
  dplyr::left_join(
    published |>
      tidyr::pivot_longer(c(auclast, cmax), names_to = "PPTESTCD",
                          values_to = "reference"),
    by = c("grp", "PPTESTCD")
  ) |>
  dplyr::mutate(
    pct_diff_at_f096 = 100 * (geomean / reference - 1),
    pct_diff_at_f1   = 100 * (at_f1 / reference - 1),
    PPTESTCD         = nlmixr2lib::ncaParamLabel(PPTESTCD)
  )

rescaled |>
  dplyr::mutate(dplyr::across(c(geomean, at_f1, reference), \(x) signif(x, 4)),
                dplyr::across(dplyr::starts_with("pct_diff"), \(x) round(x, 2)),
                cv_pct = round(cv_pct, 2)) |>
  dplyr::select(grp, PPTESTCD, reference, geomean, pct_diff_at_f096,
                at_f1, pct_diff_at_f1, cv_pct) |>
  dplyr::rename(
    "Weight band"        = grp,
    "NCA parameter"      = PPTESTCD,
    "Published"          = reference,
    "Simulated (F=0.96)" = geomean,
    "% diff (F=0.96)"    = pct_diff_at_f096,
    "Simulated (F=1.00)" = at_f1,
    "% diff (F=1.00)"    = pct_diff_at_f1,
    "Simulated CV%"      = cv_pct
  ) |>
  knitr::kable(caption = "Published Table 2 versus the packaged model at F = 0.96 and rescaled to F = 1.00.")
```

```{r f-rescale-assert}
# Every adolescent band agrees with Table 2 to better than 0.5% once F = 1,
# and is uniformly ~4% low at F = 0.96.
adolescent <- dplyr::filter(rescaled, grp != "Adult (70 kg)")
stopifnot(all(abs(adolescent$pct_diff_at_f1) < 0.5))
stopifnot(all(abs(adolescent$pct_diff_at_f096 + 4.17) < 0.6))
```

The published CV% values corroborate the reconstruction independently of the
bioavailability question, because a scale factor cannot change a coefficient of
variation. Table 2 reports AUC CV% of 3.51, 3.92 and 3.26 and Cmax CV% of 3.84,
4.28 and 3.58 for the three adolescent bands; the simulated CV% in the table
above match those to within about 0.1 percentage point, which confirms the
uniform-within-band weight distribution and the 0.75 allometric exponent.

The adult row is the one group that does not tighten under `F = 1` (about 1% low
on AUC and 3% low on Cmax). That is expected: Table 2 reports CV% of 18.70 for
adult AUC and 14.50 for adult Cmax, whereas a single typical 70 kg adult with
the reference renal function has a CV of exactly zero (verified above). The
adult column was therefore evidently simulated across the observed adult
covariate distribution rather than at a single typical value, so it is not a
like-for-like target for a typical-value replication.

Because the two candidate values cannot be separated from the paper alone -- the
printed 0.96 may be a typographical slip for 0.996 (the pivotal PK study cited
in the Introduction reported 99.6% bioavailability for this formulation and
regimen), or the simulation code may simply not have applied `F` -- the model
ships the value the paper actually prints for the parameter. Users who want to
reproduce Table 2 exactly can override it:

```{r f-override, eval = FALSE}
mod_f1 <- readModelDb("Hornik_2025_furosemide") |>
  rxode2::ini(lfdepot = log(1))
```

# Adult model with between-subject variability

The dosing simulations above exclude IIV by design. The final adult model does
carry inter-individual variability on all five structural parameters, so the
block below exercises that path and shows the two administration routes side by
side, as in the stratification of Figure 1. This is a forward simulation from
the model, not a replication of Figure 1 -- the observed concentrations that
figure plots are not public.

```{r adult-vpc}
set.seed(4242)
n_adult <- 200L

adult_subj <- tibble::tibble(id = seq_len(n_adult), WT = 70, CRCL = 86)
adult_obs_times <- seq(0, 24, by = 0.25)

adult_sc <- dplyr::bind_rows(
  dplyr::mutate(adult_subj, time = 0, amt = 30, evid = 1L, dur = 1, cmt = "depot"),
  dplyr::mutate(adult_subj, time = 1, amt = 50, evid = 1L, dur = 4, cmt = "depot"),
  tidyr::crossing(adult_subj, time = adult_obs_times) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, dur = NA_real_, cmt = "central")
) |>
  dplyr::mutate(route = "Subcutaneous 80 mg")

# Intravenous arm: 40 mg over 2 min, repeated 2 h later (Sect. 2.1).
adult_iv <- dplyr::bind_rows(
  dplyr::mutate(adult_subj, time = 0, amt = 40, evid = 1L, dur = 2 / 60, cmt = "central"),
  dplyr::mutate(adult_subj, time = 2, amt = 40, evid = 1L, dur = 2 / 60, cmt = "central"),
  tidyr::crossing(adult_subj, time = adult_obs_times) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, dur = NA_real_, cmt = "central")
) |>
  dplyr::mutate(id = id + 1000L, route = "Intravenous 40 mg x 2")

adult_events <- dplyr::bind_rows(adult_sc, adult_iv) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(!anyDuplicated(unique(adult_events[, c("id", "time", "evid")])))

adult_sim <- rxode2::rxSolve(
  mod, events = adult_events,
  keep = c("route"),
  returnType = "data.frame"
) |>
  as.data.frame()

stopifnot(dplyr::n_distinct(adult_sim$id) == 2L * n_adult)
```

```{r adult-vpc-plot, fig.cap = "Simulated adult profiles by route, with between-subject variability (counterpart of the Figure 1 stratification)."}
adult_sim |>
  dplyr::filter(!is.na(Cc), time > 0) |>
  dplyr::group_by(route, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50 * 1000)) +
  geom_ribbon(aes(ymin = Q05 * 1000, ymax = Q95 * 1000), alpha = 0.25) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~route) +
  scale_y_log10() +
  labs(
    x = "Time after first dose (h)",
    y = "Furosemide plasma concentration (ug/L)",
    title = "Adult model with between-subject variability, by route",
    caption = "Median and 5th-95th percentiles, 200 virtual adults at 70 kg and the reference renal function."
  ) +
  theme_bw()
```

The subcutaneous profile is flatter and more sustained than the intravenous
profile, which is the pharmacokinetic rationale for the On-Body Infusor
regimen described in Sect. 1.

Renal function is the model's only retained covariate. Its effect on clearance
follows Eq. 15 across the range observed in the adult study:

```{r crcl-effect}
tibble::tibble(CRCL = seq(60, 140, by = 5)) |>
  dplyr::mutate(
    cl        = 6.51 * (CRCL / 86)^0.74,
    auc_ug_h_L = 0.96 * 80 / cl * 1000
  ) |>
  ggplot(aes(CRCL, cl)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = 86, linetype = "dashed") +
  labs(
    x = "Creatinine clearance (mL/min/1.73 m^2)",
    y = "Clearance (L/h)",
    title = "Equation 15: renal function effect on clearance in a 70 kg adult",
    caption = "Dashed line marks the reference value of 86 used for normalization."
  ) +
  theme_bw()
```

# Assumptions and deviations

- **Bioavailability.** The packaged `lfdepot` is the paper's printed estimate of
  0.96 (Sect. 3.2). The paper's own Table 2 back-solves to `F = 1.00`, so every
  simulated exposure here is uniformly about 4% below the published value. The
  discrepancy is documented and quantified above rather than tuned away. Table 1
  does not report `theta(6)`, so 0.96 could not be cross-checked against the
  logit-scale estimate of Eq. 19.
- **Bioavailability parameterization.** The paper estimates `F` on the logit
  scale (Eq. 19). The model file uses the library's canonical `lfdepot`
  (log scale) at the same point estimate; the two parameterizations differ only
  in the estimation transform, not in the predicted value.
- **Weight in the adult model.** Weight was screened but **not** retained as a
  covariate in the adult fit (Sect. 3.2, Table S1). The allometric terms in the
  model file are the authors' Eqs. 9-13, which they applied post hoc to
  extrapolate the adult model to adolescents. Both size terms evaluate to 1 at
  `WT = 70`, so the model reduces exactly to the published adult model
  (Eqs. 14-19) at the reference weight. Simulating adults away from 70 kg
  applies scaling the adult fit did not itself support.
- **Covariate naming.** The paper calls the renal covariate "eGFR" and states it
  was computed with Cockcroft-Gault (Sect. 2.4, units mL/min), while the
  supplement's Figures S1-S5 label the same axis "Creatinine Clearance,
  mL/min/1.73 m^2". The register's canonical `CRCL` column is used with the
  supplement's units; the reference value of 86 is unambiguous either way.
- **Random-effects correlation.** Equation 2 defines how a CL-V correlation
  would be computed, but Table 1 reports diagonal IIV terms only and no
  off-diagonal element is published, so the etas are independent in the model
  file.
- **IIV scale.** Table 1 reports IIV as CV%. These are converted to variances as
  `omega^2 = log(CV^2 + 1)`. If the authors instead reported
  `sqrt(omega^2) * 100`, the variances would be up to 5% larger at the largest
  CV (33.1%); this does not affect any result above, because the dosing
  simulations exclude IIV entirely.
- **Cohort size.** The paper simulated 1000 virtual adolescents per weight band;
  200 per band is used here (the library cap). With IIV excluded and weight laid
  out on a deterministic grid, the group geometric means and CV% are unaffected,
  as the CV% agreement with Table 2 confirms.
- **Adult demographics.** The paper reports no baseline demographics table. The
  weight, creatinine clearance, body mass index and serum creatinine ranges
  quoted in the Population section were read from the axes of supplementary
  Figures S1-S5 and are approximate; they are recorded as figure-derived in the
  model file's `population` metadata. Only age (median 69, range 52-83) and sex
  (93.3% male) come from the text.
- **Adult reference row of Table 2.** Reported with CV% of 18.70 (AUC) and 14.50
  (Cmax), which a single typical 70 kg adult at the reference renal function
  cannot produce. That column was evidently simulated across the adult covariate
  distribution, so it is not a like-for-like typical-value target and is
  reproduced less closely than the three adolescent bands.
- **Observation grid.** The paper simulated at 0.1 h intervals over 24 h. That
  grid is used for the Table 2 comparison; it is extended to 72 h at 0.5 h
  spacing purely to support the mass-balance identity check, which needs a
  window many half-lives long.
