---
title: "Methylprednisolone IL-6 and IL-10 in neonatal cardiopulmonary bypass (Hornik 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methylprednisolone IL-6 and IL-10 in neonatal cardiopulmonary bypass (Hornik 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Hornik 2019 developed one population PK model of methylprednisolone and two
separate sequential PK/PD models, one for interleukin-6 (IL-6) and one for
interleukin-10 (IL-10). Each PK/PD model is packaged as its own file carrying
the shared PK model:

* `Hornik_2019_methylprednisolone_il6`: `r rxode2::rxode(readModelDb("Hornik_2019_methylprednisolone_il6"))$description`
* `Hornik_2019_methylprednisolone_il10`: `r rxode2::rxode(readModelDb("Hornik_2019_methylprednisolone_il10"))$description`
* Citation: `r rxode2::rxode(readModelDb("Hornik_2019_methylprednisolone_il6"))$reference`
* Article: <https://doi.org/10.1002/psp4.12470> (open access; the supplement
  holds the three NONMEM control streams, Data S1 = IL-10, Data S2 = IL-6,
  Data S3 = PK)

# Population

Sixty-four neonates (median gestational age 39 weeks, postnatal age 7 days,
postmenstrual age 40 weeks, weight 3.2 kg, range 2.2-4.3; 47% female; 58%
White, 25% Black) undergoing congenital heart surgery on cardiopulmonary bypass
(CPB; median CPB time 156.5 min, range 64-251) in the randomized trial
NCT00934843 (Table 1). All received methylprednisolone sodium succinate
30 mg/kg IV over 1 h at CPB induction; 35 of them also received a dose about
10 h before CPB. They contributed 290 methylprednisolone concentrations, 314
IL-6 observations (62 neonates) and 324 IL-10 observations (64 neonates).
RACHS-1 was below 4 in 42% and 4 or higher in 58%.

The same information is available programmatically via
`readModelDb("Hornik_2019_methylprednisolone_il6")()$population`.

# Source trace

Every `ini()` value carries an in-file comment naming its source. Summary:

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` (CL pre-CPB, 3.2 kg) | 3.88 L/h | Table 2 |
| `lvc` | 8.92 L | Table 2 |
| `lq` | 0.10 L/h | Table 2 |
| `lvp` | 16.81 L | Table 2 |
| `lka` (formation rate Kf) | 0.41 1/h | Table 2 |
| `e_wt_cl_q` | 1.24 | Table 2 |
| `e_wt_vc_vp` | 1 (not estimated) | Methods; Data S3 `LSV` |
| `e_t_cpb_cl` | -0.47 | Table 2; footnote equation |
| `etalcl`, `etalvc`, `etalq` | 47.2, 26.4, 32.6 %CV | Table 2 |
| `propSd` | 42.8% | Table 2 |
| IL-6 `limax` | 1 (fixed) | Table 3 |
| IL-6 `lic50` | 14 ng/mL | Table 3 |
| IL-6 `lrbase` | 7.9 pg/mL | Table 3 |
| IL-6 `lkout` | 0.171 1/h | Table 3 |
| IL-6 `lhill` | 2.53 | Table 3 |
| IL-6 `lcpbe` | 48.6 | Table 3 |
| IL-6 `pct_cpb_noint` (PER) | 21.4% | Table 3 |
| IL-6 `lthalf_cpb` (CPBH) | 9.08 h | Table 3 |
| IL-6 `e_rachs1_cpbe` | 2.59 | Table 3; Results |
| IL-6 `etalrbase`, `etalcpbe` | 100.5, 83.6 %CV | Table 3 |
| IL-6 `propSd_il6` | 54.1% | Table 3 |
| IL-10 `lsmax` | 2.28 | Table 4 |
| IL-10 `lsc50` | 58.2 ng/mL | Table 4 |
| IL-10 `lrbase` | 1.52 pg/mL | Table 4 |
| IL-10 `lkout` | 0.542 1/h | Table 4 |
| IL-10 `lhill` | 3.58 | Table 4 |
| IL-10 `lcpbe` | 45.7 | Table 4 |
| IL-10 `e_page_cpbe` | 14.8 | Table 4; Results |
| IL-10 `etalsmax`, `etalrbase`, `etalcpbe` | 110, 64.7, 88.1 %CV | Table 4 |
| IL-10 `propSd_il10` | 53.8% | Table 4 |
| PK ODEs (depot -> central <-> peripheral1) | n/a | Methods; Data S3 (`ADVAN4 TRANS4`, `KA` = Kf) |
| CL covariate model | n/a | Table 2 footnote; Data S3 `TVCL` |
| IL-6 ODE (partial interaction) | n/a | Eqs. 3-5; Data S2 `$DES` |
| CPB time course `cpbv` | n/a | Eq. 6; Data S2 `CPBV`, `CPBES` |
| IL-10 ODE (complete interaction, Smax) | n/a | Methods text after Eq. 6; Data S1 `$DES` |

# CPB event tables

The model reads the CPB time course from two time-varying indicators,
`CPB_ON` (1 while on bypass) and `CPB_POST` (1 after coming off bypass), plus
the subject's total bypass time `T_CPB` (min). The helper below builds one
subject's records: a dense observation grid, extra records at the CPB start
and end so the indicators switch at the right times, and the prodrug doses
into `depot` as 1 h infusions. The first observation is on `il6` or `il10`
(the PD state) because each model has two endpoints and one of them is an
ODE state.

Hornik 2019 simulated a dose at CPB initiation, or that dose plus a dose 8 h
earlier (Methods, Dosing simulation). CPB starts at 8 h here so both regimens
share one time axis, and AUC0-24 is taken from CPB start (Eq. 7).

```{r helpers}
t_cpb_start <- 8
obs_grid <- seq(0, t_cpb_start + 24, by = 0.25)

make_subject_events <- function(id, dose_mgkg, n_doses, WT, T_CPB, obs_cmt) {
  t_end <- t_cpb_start + T_CPB / 60
  times <- sort(unique(c(obs_grid, t_cpb_start, t_end)))
  obs <- data.frame(id = id, time = times, evid = 0, amt = 0, rate = 0, cmt = obs_cmt)
  if (dose_mgkg > 0) {
    dose_times <- if (n_doses == 2) c(t_cpb_start - 8, t_cpb_start) else t_cpb_start
    amt <- dose_mgkg * WT
    doses <- data.frame(id = id, time = dose_times, evid = 1, amt = amt, rate = amt, cmt = "depot")
    obs <- dplyr::bind_rows(obs, doses)
  }
  obs |>
    dplyr::arrange(time, dplyr::desc(evid)) |>
    dplyr::mutate(
      WT = WT, T_CPB = T_CPB,
      CPB_ON = as.integer(time >= t_cpb_start & time < t_end),
      CPB_POST = as.integer(time >= t_end)
    )
}

auc_from_cpb <- function(time, y) {
  keep <- time >= t_cpb_start & time <= t_cpb_start + 24
  t <- time[keep]
  v <- y[keep]
  sum(diff(t) * (head(v, -1) + tail(v, -1)) / 2)
}

regimens <- tibble::tribble(
  ~regimen,         ~dose_mgkg, ~n_doses,
  "Placebo",        0,          1,
  "10 mg/kg x 1",   10,         1,
  "10 mg/kg x 2",   10,         2,
  "30 mg/kg x 1",   30,         1,
  "30 mg/kg x 2",   30,         2
)
```

# Typical-value check against the published regimen ratios

The paper states that simulated IL-6 was more than 50% lower and IL-10 more than
100% higher after methylprednisolone than after placebo, with little difference
between 10 and 30 mg/kg or between one and two doses. With the random effects
zeroed, a 3.2 kg neonate with RACHS-1 below 4, PMA 40 weeks and a
156.5 min bypass gives:

```{r typical}
mod6 <- readModelDb("Hornik_2019_methylprednisolone_il6")
mod10 <- readModelDb("Hornik_2019_methylprednisolone_il10")
typ6 <- rxode2::zeroRe(mod6)
typ10 <- rxode2::zeroRe(mod10)

typical_auc <- function(mod, endpoint, extra) {
  res <- lapply(seq_len(nrow(regimens)), function(i) {
    ev <- make_subject_events(1, regimens$dose_mgkg[i], regimens$n_doses[i],
      WT = 3.2, T_CPB = 156.5, obs_cmt = endpoint
    )
    for (nm in names(extra)) ev[[nm]] <- extra[[nm]]
    s <- as.data.frame(rxode2::rxSolve(mod, ev, returnType = "data.frame"))
    data.frame(regimen = regimens$regimen[i], auc = auc_from_cpb(s$time, s[[endpoint]]))
  })
  dplyr::bind_rows(res)
}

typ_auc <- dplyr::bind_rows(
  typical_auc(typ6, "il6", list(RACHS1 = 3)) |> dplyr::mutate(endpoint = "IL-6"),
  typical_auc(typ10, "il10", list(PAGE = 40)) |> dplyr::mutate(endpoint = "IL-10")
)
typ_wide <- tidyr::pivot_wider(typ_auc, names_from = regimen, values_from = auc)
typ_ratio <- typ_wide |>
  dplyr::transmute(
    endpoint,
    `10 mg/kg vs placebo` = `10 mg/kg x 1` / Placebo,
    `30 vs 10 mg/kg` = `30 mg/kg x 1` / `10 mg/kg x 1`,
    `2 vs 1 dose` = `10 mg/kg x 2` / `10 mg/kg x 1`
  )
knitr::kable(typ_ratio, digits = 3, caption = "Typical-subject AUC0-24 ratios.")
```

IL-6 10 mg/kg versus placebo is `r round(typ_ratio[["10 mg/kg vs placebo"]][1], 2)`
(Table 5 mean ratio 0.27). This is the check on concentration units: the
control streams compute the PD driver as `A(2)/V2` without the 1000 scaling
used for the PK observations, so IC50 and SC50 could be read as mg/L. With
IC50 = 14 mg/L the drug effect would be negligible at these concentrations
(peak about 3.4 mg/L after 30 mg/kg) and the ratio would be close to 1. The
paper's ng/mL units are the only reading that reproduces Table 5.

```{r typical-assert}
stopifnot(
  abs(typ_ratio[["10 mg/kg vs placebo"]][1] / 0.27 - 1) < 0.10,
  typ_ratio[["30 vs 10 mg/kg"]][1] < 1,
  typ_ratio[["30 vs 10 mg/kg"]][2] > 1,
  typ_ratio[["10 mg/kg vs placebo"]][2] > 2
)
```

# Virtual cohorts

Hornik 2019 generated 1,000 term infants with PK-Sim, which is not reproducible
here. The cohorts below use 200 virtual neonates per endpoint. Each neonate
receives all five regimens with the same random effects, so ratios are paired
the way the paper's Table 5 ratios are. Weights are drawn from a normal
distribution matching Table 1 (median 3.2 kg, truncated to 2.2-4.3 kg).
Postnatal age is uniform on 0-28 days and PMA = 40 weeks + PNA (term
infants). For IL-6, RACHS-1 is below 4 or 4 or higher with equal probability;
CPB time is uniform on 60-240 min for both endpoints.

```{r cohort}
set.seed(20190101)
n_sub <- 200

draw_cohort <- function(n) {
  tibble::tibble(
    base_id = seq_len(n),
    WT = pmin(pmax(rnorm(n, 3.2, 0.45), 2.2), 4.3),
    PNA = runif(n, 0, 28),
    PAGE = 40 + PNA / 7,
    RACHS1 = sample(c(3, 4), n, replace = TRUE),
    T_CPB = runif(n, 60, 240)
  )
}

draw_etas <- function(mod, n) {
  # Every omega in both models is diagonal (Tables 2-4 report no
  # covariances), so independent normal draws are exact.
  om <- rxode2::rxode2(mod)$omega
  stopifnot(all(om[upper.tri(om)] == 0))
  e <- vapply(sqrt(diag(om)), function(s) rnorm(n, 0, s), numeric(n))
  as.data.frame(matrix(e, nrow = n, dimnames = list(NULL, rownames(om))))
}

build_events <- function(cohort, etas, endpoint, covs) {
  out <- vector("list", nrow(regimens) * nrow(cohort))
  k <- 0
  for (r in seq_len(nrow(regimens))) {
    for (i in seq_len(nrow(cohort))) {
      k <- k + 1
      ev <- make_subject_events((r - 1) * nrow(cohort) + i,
        regimens$dose_mgkg[r], regimens$n_doses[r],
        WT = cohort$WT[i], T_CPB = cohort$T_CPB[i], obs_cmt = endpoint
      )
      for (nm in covs) ev[[nm]] <- cohort[[nm]][i]
      for (nm in names(etas)) ev[[nm]] <- etas[[nm]][i]
      ev$base_id <- cohort$base_id[i]
      ev$regimen <- regimens$regimen[r]
      out[[k]] <- ev
    }
  }
  dplyr::bind_rows(out)
}

cohort6 <- draw_cohort(n_sub)
cohort10 <- draw_cohort(n_sub)
ev6 <- build_events(cohort6, draw_etas(mod6, n_sub), "il6", "RACHS1")
ev10 <- build_events(cohort10, draw_etas(mod10, n_sub), "il10", "PAGE")
stopifnot(
  length(unique(ev6$id)) == n_sub * nrow(regimens),
  length(unique(ev10$id)) == n_sub * nrow(regimens)
)
```

# Simulation

The random effects are drawn above and passed in as data columns, so the
solves use the zero-random-effect models; residual error is not added
because Table 5 and Tables S4-S5 summarise model-predicted profiles.

```{r simulate}
sim6 <- as.data.frame(rxode2::rxSolve(typ6, ev6,
  keep = c("base_id", "regimen", "RACHS1", "WT"), returnType = "data.frame"
))
sim10 <- as.data.frame(rxode2::rxSolve(typ10, ev10,
  keep = c("base_id", "regimen", "PAGE", "WT"), returnType = "data.frame"
))
stopifnot(!anyNA(sim6$il6), !anyNA(sim10$il10))
```

# Replicate published figures

```{r figure-1}
# Replicates Figure 1 of Hornik 2019: simulated IL-6 by regimen.
pi_plot <- function(sim, endpoint, ylab, title) {
  sim |>
    dplyr::group_by(time, regimen) |>
    dplyr::summarise(
      Q05 = quantile(.data[[endpoint]], 0.05),
      Q50 = quantile(.data[[endpoint]], 0.50),
      Q95 = quantile(.data[[endpoint]], 0.95),
      .groups = "drop"
    ) |>
    ggplot(aes(time - t_cpb_start, Q50, colour = regimen, fill = regimen)) +
    geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.1, colour = NA) +
    geom_line() +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(x = "Time after CPB start (h)", y = ylab, title = title)
}
pi_plot(sim6, "il6", "IL-6 (pg/mL)", "Simulated IL-6, median and 90% interval") +
  labs(caption = "Replicates Figure 1 of Hornik 2019.")
```

```{r figure-2}
# Replicates Figure 2 of Hornik 2019: simulated IL-10 by regimen.
pi_plot(sim10, "il10", "IL-10 (pg/mL)", "Simulated IL-10, median and 90% interval") +
  labs(caption = "Replicates Figure 2 of Hornik 2019.")
```

# AUC0-24 after CPB start (Table 5, Tables S4-S5)

```{r auc}
auc6 <- sim6 |>
  dplyr::group_by(base_id, regimen, RACHS1) |>
  dplyr::summarise(auc = auc_from_cpb(time, il6), .groups = "drop")
auc10 <- sim10 |>
  dplyr::group_by(base_id, regimen, PAGE) |>
  dplyr::summarise(auc = auc_from_cpb(time, il10), .groups = "drop")

paired_ratios <- function(auc) {
  w <- tidyr::pivot_wider(auc, names_from = regimen, values_from = auc)
  r <- list(
    `2 vs 1 dose` = c(w$`10 mg/kg x 2` / w$`10 mg/kg x 1`, w$`30 mg/kg x 2` / w$`30 mg/kg x 1`),
    `30 vs 10 mg/kg` = c(w$`30 mg/kg x 1` / w$`10 mg/kg x 1`, w$`30 mg/kg x 2` / w$`10 mg/kg x 2`),
    `10 mg/kg vs placebo` = c(w$`10 mg/kg x 1` / w$Placebo, w$`10 mg/kg x 2` / w$Placebo)
  )
  tibble::tibble(
    ratio = names(r),
    sim_mean = vapply(r, mean, 1),
    sim_median = vapply(r, median, 1)
  )
}
published_ratio <- tibble::tribble(
  ~ratio,                ~pub_il6, ~pub_il10,
  "2 vs 1 dose",         0.92,     1.08,
  "30 vs 10 mg/kg",      0.89,     1.03,
  "10 mg/kg vs placebo", 0.27,     3.85
)
r6 <- paired_ratios(auc6)
r10 <- paired_ratios(auc10)
ratio_tab <- published_ratio |>
  dplyr::left_join(dplyr::rename(r6, il6_mean = sim_mean, il6_median = sim_median), by = "ratio") |>
  dplyr::left_join(dplyr::rename(r10, il10_mean = sim_mean, il10_median = sim_median), by = "ratio")
ratio_tab |>
  dplyr::select(ratio, pub_il6, il6_mean, il6_median, pub_il10, il10_mean, il10_median) |>
  dplyr::rename(
    "Ratio" = ratio,
    "IL-6 Table 5 mean" = pub_il6,
    "IL-6 simulated mean" = il6_mean,
    "IL-6 simulated median" = il6_median,
    "IL-10 Table 5 mean" = pub_il10,
    "IL-10 simulated mean" = il10_mean,
    "IL-10 simulated median" = il10_median
  ) |>
  knitr::kable(digits = 2, caption = "Paired AUC0-24 ratios versus Hornik 2019 Table 5.")
```

The IL-6 ratios and the IL-10 dose-level and dose-number ratios reproduce
Table 5 closely. The IL-10 methylprednisolone-versus-placebo ratio is a
mean of per-subject ratios driven by the 110 %CV on Smax, so it is heavy-tailed
and sensitive to the cohort. Its median sits below the published mean, as a
right-skewed ratio's median should.

```{r auc-strata}
s4 <- auc6 |>
  dplyr::mutate(RACHS = ifelse(RACHS1 >= 4, ">=4", "<4")) |>
  dplyr::group_by(RACHS, regimen) |>
  dplyr::summarise(sim_median = median(auc), .groups = "drop")
# Table S4 medians averaged over the three CPB-time strata, 1 dose rows
# (placebo 1-dose rows) -- RACHS-1 < 4: 30 mg/kg 909, 10 mg/kg 1010,
# placebo 3388; RACHS-1 >= 4: 30 mg/kg 2146, 10 mg/kg 2411, placebo 6572.
s4_pub <- tibble::tribble(
  ~RACHS, ~regimen,       ~pub_median,
  "<4",   "30 mg/kg x 1", mean(c(870, 910, 948)),
  "<4",   "10 mg/kg x 1", mean(c(988, 1004, 1037)),
  "<4",   "Placebo",      mean(c(3172, 3403, 3590)),
  ">=4",  "30 mg/kg x 1", mean(c(2056, 2136, 2246)),
  ">=4",  "10 mg/kg x 1", mean(c(2362, 2393, 2479)),
  ">=4",  "Placebo",      mean(c(6375, 6617, 6725))
)
s4_cmp <- dplyr::inner_join(s4_pub, s4, by = c("RACHS", "regimen")) |>
  dplyr::mutate(pct_diff = 100 * (sim_median / pub_median - 1))
s4_cmp |>
  dplyr::rename(
    "RACHS-1" = RACHS, "Regimen" = regimen,
    "Table S4 median" = pub_median, "Simulated median" = sim_median,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = 0, caption = "IL-6 AUC0-24 (pg*h/mL) by RACHS-1 stratum versus Table S4.")
```

```{r auc-strata-il10}
s5 <- auc10 |>
  dplyr::mutate(PMA = cut(PAGE, c(-Inf, 41, 42, Inf), labels = c("<=41", "41.1-42", ">42"))) |>
  dplyr::group_by(PMA, regimen) |>
  dplyr::summarise(sim_median = median(auc), .groups = "drop")
# Table S5 medians averaged over the three CPB-time strata, 1-dose rows.
s5_pub <- tibble::tribble(
  ~PMA,      ~regimen,       ~pub_median,
  "<=41",    "30 mg/kg x 1", mean(c(454, 698, 930)),
  "<=41",    "10 mg/kg x 1", mean(c(422, 674, 913)),
  "<=41",    "Placebo",      mean(c(165, 252, 337)),
  "41.1-42", "30 mg/kg x 1", mean(c(595, 943, 1255)),
  "41.1-42", "10 mg/kg x 1", mean(c(559, 903, 1227)),
  "41.1-42", "Placebo",      mean(c(222, 354, 475)),
  ">42",     "30 mg/kg x 1", mean(c(879, 1357, 1838)),
  ">42",     "10 mg/kg x 1", mean(c(835, 1306, 1783)),
  ">42",     "Placebo",      mean(c(339, 550, 754))
)
s5_cmp <- s5 |>
  dplyr::mutate(PMA = as.character(PMA)) |>
  dplyr::inner_join(s5_pub, by = c("PMA", "regimen")) |>
  dplyr::mutate(pct_diff = 100 * (sim_median / pub_median - 1))
s5_cmp |>
  dplyr::select(PMA, regimen, pub_median, sim_median, pct_diff) |>
  dplyr::rename(
    "PMA (weeks)" = PMA, "Regimen" = regimen,
    "Table S5 median" = pub_median, "Simulated median" = sim_median,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = 0, caption = "IL-10 AUC0-24 (pg*h/mL) by PMA stratum versus Table S5.")
```

Most IL-10 stratum medians fall within about 20% of Table S5. The
least-mature placebo stratum is lower, which the uniform-PNA cohort (PMA
40-41 weeks within that stratum) and the steep PMA exponent of 14.8
explain. The IL-6 medians run 30-75% above Table S4, placebo rows most. The gap is already present
for the typical subject: RACHS-1 below 4, 156.5 min bypass, placebo gives
`r round(typ_wide$Placebo[1])` pg*h/mL against about 3400 in the
120-180 min row of Table S4. Yet the IL-6 regimen ratios reproduce Table 5. A
difference that scales every IL-6 exposure about equally points to how the
paper's NONMEM simulation evaluated the CPB time course, not to a
transcription error. The CPB effect enters every regimen, placebo
included, while the drug parameters would move the ratios. NONMEM evaluates
`$PK` quantities such as `CPBV`, `CPBES` and the withdrawal term
`EXP(-0.693*TACPB/CPBH)` once per data record, so they are piecewise
constant over the simulation grid. The simulation data set is not
published, so this cannot be checked. The parameters are not adjusted to
close the gap.

```{r auc-assert}
stopifnot(
  # Structural: a wrong IC50 unit, Hill or CPB term moves these by far more.
  abs(r6$sim_mean[r6$ratio == "10 mg/kg vs placebo"] / 0.27 - 1) < 0.25,
  abs(r6$sim_mean[r6$ratio == "30 vs 10 mg/kg"] / 0.89 - 1) < 0.10,
  abs(r10$sim_median[r10$ratio == "30 vs 10 mg/kg"] / 1.03 - 1) < 0.10,
  r10$sim_median[r10$ratio == "10 mg/kg vs placebo"] > 2,
  # IL-10 strata: centre and envelope (each median has ~15% Monte-Carlo SE).
  abs(median(s5_cmp$pct_diff)) < 25,
  quantile(abs(s5_cmp$pct_diff), 0.9) < 50,
  # IL-6 strata: the documented level gap above, bounded at a factor of 2.5
  # per row so that a mis-transcribed CPBE or baseline (a multi-fold shift)
  # still fails.
  all(abs(log(s4_cmp$sim_median / s4_cmp$pub_median)) < log(2.5))
)
```

# PKNCA validation of the methylprednisolone PK

The paper reports no NCA, but it states typical values for a 3.2 kg neonate
(CL/F 3.88 L/h, Table 2; the Discussion rounds to 3.8 L/h) and median post hoc
clearances of 1.28 L/h/kg before CPB and 1.24 L/h/kg after CPB. For a typical
3.2 kg neonate with a 156.5 min bypass, pre- and post-CPB clearance are
identical, so the PKNCA clearance from a single 30 mg/kg dose must return
3.88 L/h. The half-life should match the slower eigenvalue of the
two-compartment system.

```{r pknca-typical}
ev_pk <- make_subject_events(1, 30, 1, WT = 3.2, T_CPB = 156.5, obs_cmt = "il6") |>
  dplyr::mutate(RACHS1 = 3)
grid_long <- sort(unique(c(ev_pk$time[ev_pk$evid == 0], seq(32, 400, by = 2))))
ev_pk <- dplyr::bind_rows(
  ev_pk,
  data.frame(id = 1, time = setdiff(grid_long, ev_pk$time), evid = 0, amt = 0, rate = 0, cmt = "il6") |>
    dplyr::mutate(WT = 3.2, T_CPB = 156.5, CPB_ON = 0, CPB_POST = 1, RACHS1 = 3)
) |>
  dplyr::arrange(time, dplyr::desc(evid))
ev_pk$treatment <- "30 mg/kg x 1"
sim_pk <- as.data.frame(rxode2::rxSolve(typ6, ev_pk,
  returnType = "data.frame", rtol = 1e-10, atol = 1e-12
))
sim_pk$treatment <- "30 mg/kg x 1"
sim_pk$id <- 1L # a single-subject solve returns no id column

conc_pk <- sim_pk |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(time = time - t_cpb_start, conc = pmax(Cc, 0) / 1000) |>
  dplyr::filter(time >= 0) |>
  dplyr::select(id, time, conc, treatment)
conc_pk <- dplyr::bind_rows(conc_pk, dplyr::distinct(conc_pk, id, treatment) |> dplyr::mutate(time = 0, conc = 0)) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)
dose_pk <- ev_pk |>
  dplyr::filter(evid == 1) |>
  dplyr::mutate(time = time - t_cpb_start) |>
  dplyr::select(id, time, amt, treatment)

nca_pk <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_pk, conc ~ time | treatment + id),
  PKNCA::PKNCAdose(dose_pk, amt ~ time | treatment + id),
  intervals = data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                         aucinf.obs = TRUE, half.life = TRUE, cl.obs = TRUE)
))

vc <- 8.92; vp <- 16.81; cl <- 3.88; q <- 0.10
k10 <- cl / vc; k12 <- q / vc; k21 <- q / vp
beta <- ((k10 + k12 + k21) - sqrt((k10 + k12 + k21)^2 - 4 * k10 * k21)) / 2
published_pk <- tibble::tibble(treatment = "30 mg/kg x 1", cl.obs = 3.88, half.life = log(2) / beta)
cmp_pk <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_pk,
  reference = published_pk,
  by = "treatment",
  units = c(cl.obs = "L/h", half.life = "h"),
  tolerance_pct = 20
)
knitr::kable(cmp_pk, caption = "Typical 3.2 kg neonate: PKNCA versus Table 2 CL and the model's terminal half-life. * differs by >20%.")
nca_tab <- as.data.frame(nca_pk$result)
stopifnot(
  abs(nca_tab$PPORRES[nca_tab$PPTESTCD == "cl.obs"] / 3.88 - 1) < 0.02,
  abs(nca_tab$PPORRES[nca_tab$PPTESTCD == "half.life"] / (log(2) / beta) - 1) < 0.10
)
```

The long terminal half-life (`r round(log(2) / beta)` h) follows from the small
intercompartmental clearance (Q = 0.10 L/h) against a large peripheral
volume. It is a property of the published parameters and is invisible over
the study's 24 h post-CPB sampling window.

Per-kilogram individual clearance in the stochastic IL-6 cohort, 30 mg/kg
single-dose regimen, compared with the paper's median post hoc estimates:

```{r cl-per-kg}
cl_kg <- sim6 |>
  dplyr::filter(regimen == "30 mg/kg x 1") |>
  dplyr::mutate(phase = ifelse(time < t_cpb_start, "pre-CPB", ifelse(time > t_cpb_start + 4.5, "post-CPB", NA))) |>
  dplyr::filter(!is.na(phase)) |>
  dplyr::group_by(base_id, phase) |>
  dplyr::summarise(cl_kg = dplyr::first(cl) / dplyr::first(WT), .groups = "drop") |>
  dplyr::group_by(phase) |>
  dplyr::summarise(sim_median = median(cl_kg), .groups = "drop") |>
  dplyr::left_join(tibble::tibble(phase = c("pre-CPB", "post-CPB"), published = c(1.28, 1.24)), by = "phase")
cl_kg |>
  dplyr::rename("Phase" = phase, "Simulated median (L/h/kg)" = sim_median, "Published EBE median (L/h/kg)" = published) |>
  knitr::kable(digits = 2)
stopifnot(all(abs(cl_kg$sim_median / cl_kg$published - 1) < 0.25))
```

# Assumptions and deviations

- **CPB timing encoded as time-varying indicators.** The control streams
  derive the CPB windows from a sample index (`POINT`) and a `CPBSTART` data
  column. Here `CPB_ON` and `CPB_POST` are data columns and the 0.5 h onset
  delay (`CPBES`) and post-CPB withdrawal of the IL-6 CPB effect are computed
  inside the model from two bookkeeping states (`cpb_elapsed`, `cpb_decay`),
  which reproduces Eq. 6 exactly in continuous time.
- **POSTCPB.** The control streams code the clearance switch as
  `STA4 = STA2 + STA3`, whose flag definitions are not given. The paper names
  the indicator POSTCPB ("time after CPB") and reports pre- versus post-CPB
  clearance, so it is encoded as `CPB_POST`.
- **PK values.** The PD control streams (Data S1, S2) carry PK thetas that
  differ from Table 2 (V2 8.848, V3 11.58, Q 0.0868 versus 8.92, 16.81,
  0.10). The PD fits used individual PK estimates, so these thetas did not
  drive the PD. The published final estimates in Table 2 are used. The
  Discussion's Vss/F of 26.3 L is closer to Table 2 (Vc + Vp = 25.7 L) than
  to the stream values (20.4 L).
- **IIV scale.** Tables 2-4 report IIV as %CV only, and the control-stream
  `$OMEGA` initials are not near-converged, so they cannot settle the scale.
  The library convention omega^2 = log(1 + CV^2) is used. For the large IL-6
  and IL-10 PD variabilities (83.6-110 %CV) the alternative reading
  omega^2 = CV^2 would give wider distributions.
- **Residual error.** NONMEM `Y = EFF*EXP(ERR(1))` under FOCE-I is
  linearised to a proportional error, and the paper reports it as
  "Proportional error, %". It is encoded as `prop()` with SD = %/100.
- **PD driver units.** The streams use `A(2)/V2` (mg/L) inside `$DES`, while
  the paper reports IC50 and SC50 in ng/mL. The model uses the ng/mL `Cc`,
  and the typical-value check above shows this is the only reading that
  reproduces Table 5.
- **Dosing route.** As in the paper, every prodrug dose (including a dose
  into the CPB circuit) enters one depot and converts completely to
  methylprednisolone at rate Kf, with no prodrug renal clearance
  (Discussion, limitations).
- **IL-6 absolute exposure versus Table S4.** Simulated IL-6 AUC0-24 medians
  are 30-75% higher than Table S4 in every stratum, while the Table 5
  regimen ratios, the IL-10 Table S5 medians and the PK typical values
  reproduce. The model is not adjusted; see the discussion under the
  AUC tables for the likely cause (record-wise evaluation of the CPB time
  course in the paper's NONMEM simulation).
- **Virtual population.** The PK-Sim cohort of the paper (1,000 term
  infants, 50% female, 85% White) is replaced by the simple normal-weight,
  uniform-PNA cohort above. The Table S4/S5 comparison is therefore
  approximate; the paired ratios in Table 5 depend much less on the
  covariate distribution.
