---
title: "Growth curves in youth-onset type 2 diabetes (Hosey 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Growth curves in youth-onset type 2 diabetes (Hosey 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Hosey_2022_t2dm_growth"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Article (open access, PMC9010268): <https://doi.org/10.1111/cts.13207>\n")
```

Hosey 2022 is not a drug model. It supplies the **anthropometric system layer**
for pediatric physiologically based pharmacokinetic (PBPK) models of youth with
type 2 diabetes mellitus (T2DM). PBPK scaling factors such as liver volume are
functions of body surface area, and so of height and weight. The standard CDC
growth charts describe typically developing children and stop at the 97th
percentile, which leaves most children with T2DM outside them. The authors
therefore refitted the CDC functional forms to the heights and weights of
children who developed T2DM:

* **Height** uses the CDC three-logistic stature function
  $$
  \mathrm{Ht}(\mathrm{age}) = \frac{a q}{1 + e^{-b_1(\mathrm{age}-c_1)}}
  + \frac{a p}{1 + e^{-b_2(\mathrm{age}-c_2)}}
  + \frac{f - a}{1 + e^{-b_3(\mathrm{age}-c_3)}}
  $$
* **Weight** uses the CDC polynomial
  $\mathrm{Wt}(\mathrm{age}) = a\,\mathrm{age}^{10} + b\,\mathrm{age}^9 + \dots + j\,\mathrm{age} + k$,
  where the $\mathrm{age}^{10}$ term is present for males only.

Both equations are fitted separately for each sex, with age in years. In the
model, the rxode2 `time` variable **is** chronological age in years. The model
has two outputs, `ht` (cm) and `bw` (kg). Each carries a proportional residual
error equal to the sex- and output-specific coefficient of variation of the
held-out validation set (Table 2). These are the SD bands that Figure 2 draws
around each curve.

# Population

```{r population, echo = FALSE, results = "asis"}
pop <- ui$population
for (nm in names(pop)) cat("* **", nm, "**: ", paste(pop[[nm]], collapse = "; "), "\n", sep = "")
```

The data are 356 children (211 female, 145 male) with an ICD-9 / ICD-10
diagnosis of T2DM seen at Children's Mercy Kansas City (Table 1). Children with
type 1 diabetes were excluded. Visits before and after the T2DM diagnosis were
both used, so the curves describe the growth of children *who develop* T2DM, not
only growth after diagnosis. The target range is 2 to 18 years. Data from 19 to
25 years were added only to keep the curves from dropping off at the upper end
(Methods, Dataset). Supplemental Information S4 gives the self-identified race
and ethnicity distribution.

# Source trace

Each `ini()` entry in `inst/modeldb/endogenous/Hosey_2022_t2dm_growth.R` has an
in-file comment giving its origin. The table below collects them in one place.

| Model component | Source location |
|---|---|
| Height equation form (`ht_male`, `ht_female`) | Methods, "Model development" (CDC stature function) |
| `ht_a_male` ... `ht_c3_male` (166, 0.283, 0.717, 5.18, 1.44, 0.245, 5.11, 178, 1.8, 12.4) | Results, "Male height (cm, age in years)" equation |
| `ht_a_female` ... `ht_c3_female` (160, 0.396, 0.604, 0.537, 2.23, 0.0039, 1.77, 209, 0.473, 9.54) | Results, "Female height (cm, age in years)" equation |
| Weight equation form (`bw_male`, `bw_female`) | Methods, "Model development" (CDC weight polynomial) |
| `bw_a_male` ... `bw_k_male` | Supplemental Information S1, column "Male" (six significant figures) |
| `bw_b_female` ... `bw_k_female` | Supplemental Information S1, column "Female" (six significant figures; no $\mathrm{age}^{10}$ term) |
| `propSd_ht_male` 0.0587, `propSd_ht_female` 0.0675 | Table 2, "Coefficients of variation (%)", Height, column "Validation set fit to growth model: T2DM" |
| `propSd_bw_male` 0.378, `propSd_bw_female` 0.459 | Table 2, "Coefficients of variation (%)", Weight, same column |
| Proportional form of the residual error | Figure 2 caption (SD bands "derived from the coefficient of variation (%) of the datapoints from the model"); checked numerically below |
| Sex selection via `SEXF` | Separate male / female equations throughout Results and S1 |

## Units

| Symbol | Units | Note |
|---|---|---|
| `time` / `age` | year | the rxode2 time variable **is** chronological age |
| `ht`, `ht_male`, `ht_female` | cm | standing height (stature) |
| `bw`, `bw_male`, `bw_female` | kg | |
| `ht_a_*`, `ht_f_*` | cm | |
| `ht_q_*`, `ht_p_*` | fraction | $p + q = 1$ for both sexes as printed |
| `ht_b1_*`, `ht_b2_*`, `ht_b3_*` | 1/year | |
| `ht_c1_*`, `ht_c2_*`, `ht_c3_*` | year | |
| `bw_<x>_*` for the coefficient of $\mathrm{age}^n$ | kg/year^n | |
| `propSd_*` | fraction | |

# Parameter table

```{r param-table, echo = FALSE}
ui$iniDf |>
  dplyr::filter(!is.na(.data$ntheta)) |>
  dplyr::transmute(
    Parameter = .data$name,
    Value = signif(.data$est, 6),
    Fixed = .data$fix,
    Label = .data$label
  ) |>
  dplyr::rename("Fixed in source" = "Fixed") |>
  knitr::kable()
```

All coefficients are the paper's final point estimates from robust nonlinear
regression, reported without uncertainty. The CVs are summary statistics of the
validation set. Every value is therefore `fixed()`.

# Simulation

## Typical curves

The typical-value outputs `ht` and `bw` do not depend on the residual error, so
they are read directly from the solved columns. One "subject" per sex is solved
on a fine age grid.

```{r simulate}
mod <- readModelDb("Hosey_2022_t2dm_growth")

solve_sex <- function(sexf, ages) {
  ev <- data.frame(id = 1L, time = ages, evid = 0L, dvid = 1L, SEXF = sexf)
  out <- rxode2::rxSolve(mod, ev, returnType = "data.frame")
  out$SEXF <- sexf
  out$sex <- ifelse(sexf == 1, "Female", "Male")
  out
}

ages_fine <- seq(2, 25, by = 0.01)
typ <- dplyr::bind_rows(solve_sex(1, ages_fine), solve_sex(0, ages_fine))
nrow(typ)
```

## Stochastic cohort

A cross-sectional virtual cohort has 200 children per sex at each of seven ages.
Each row draws a height (`dvid = 1`) and a weight (`dvid = 2`) with the
proportional residual error. This is how the paper's Supplemental Information S3
Simcyp simulation used the equations and the CVs.

```{r stochastic}
rxode2::rxSetSeed(20220101)
ages_vpc <- c(2, 5, 8, 11, 14, 16, 18)
n_per_sex <- 200
cohort <- expand.grid(
  id = seq_len(2 * n_per_sex),
  time = ages_vpc,
  dvid = 1:2
) |>
  dplyr::mutate(
    evid = 0L,
    SEXF = as.integer(.data$id > n_per_sex)
  ) |>
  dplyr::arrange(.data$id, .data$time, .data$dvid)

vpc <- rxode2::rxSolve(mod, cohort, keep = "SEXF", returnType = "data.frame") |>
  dplyr::mutate(
    output = ifelse(.data$CMT == 1, "Height (cm)", "Weight (kg)"),
    pred = ifelse(.data$CMT == 1, .data$ht, .data$bw),
    sex = ifelse(.data$SEXF == 1, "Female", "Male")
  )
nrow(vpc)
```

# Validation

## 1. The model returns the published equations

The outputs must equal the printed equations, evaluated by hand here from the
paper's numbers, to machine precision. Both sides use the same coefficients, so
any difference is transcription or algebra error and the bound is tight.

```{r exact-check}
ht_eq <- function(age, a, q, p, b1, c1, b2, c2, f, b3, c3) {
  a * q / (1 + exp(-b1 * (age - c1))) +
    a * p / (1 + exp(-b2 * (age - c2))) +
    (f - a) / (1 + exp(-b3 * (age - c3)))
}
wt_male_s1 <- c(3.44781e-08, -4.57950e-06, 2.60885e-04, -8.35633e-03, 1.65867e-01,
                -2.12088, 17.5837, -92.5981, 293.823, -499.980, 358.440)
wt_female_s1 <- c(0, -4.81428e-07, 5.00192e-05, -2.19327e-03, 5.29444e-02,
                  -0.770305, 6.95625, -38.7195, 127.524, -221.140, 164.303)
wt_eq <- function(age, coefs) vapply(age, function(x) sum(coefs * x^(10:0)), numeric(1))

chk_ages <- c(2, 4, 6, 8, 10, 12, 14, 16, 18)
chk <- typ |>
  dplyr::filter(round(.data$time, 2) %in% chk_ages) |>
  dplyr::mutate(
    ht_hand = ifelse(
      .data$SEXF == 1,
      ht_eq(.data$time, 160, 0.396, 0.604, 0.537, 2.23, 0.0039, 1.77, 209, 0.473, 9.54),
      ht_eq(.data$time, 166, 0.283, 0.717, 5.18, 1.44, 0.245, 5.11, 178, 1.8, 12.4)
    ),
    bw_hand = ifelse(
      .data$SEXF == 1,
      wt_eq(.data$time, wt_female_s1),
      wt_eq(.data$time, wt_male_s1)
    )
  )
stopifnot(
  nrow(chk) == 2 * length(chk_ages),
  all(abs(chk$ht - chk$ht_hand) < 1e-8),
  all(abs(chk$bw - chk$bw_hand) < 1e-8)
)
chk |>
  dplyr::select("sex", "time", "ht", "bw") |>
  dplyr::rename("Sex" = "sex", "Age (y)" = "time", "Height (cm)" = "ht", "Weight (kg)" = "bw") |>
  knitr::kable(digits = 1)
```

## 2. Height velocity, replicating Figure 5

The Methods define height velocity as the first derivative of the final height
models. Results reports a local peak of height velocity at **8.5 years for
females** and **12.5 years for males**. The paper binned its results in 6-month
intervals, so the peak is compared at that resolution.

```{r velocity}
vel <- typ |>
  dplyr::group_by(.data$sex) |>
  dplyr::arrange(.data$time, .by_group = TRUE) |>
  dplyr::mutate(velocity = (dplyr::lead(.data$ht) - dplyr::lag(.data$ht)) /
    (dplyr::lead(.data$time) - dplyr::lag(.data$time))) |>
  dplyr::ungroup() |>
  dplyr::filter(!is.na(.data$velocity))

# A LOCAL maximum: velocity above both neighbours. The female curve is
# still falling from its early-childhood value at 6 years, so a plain
# window maximum would return the window edge instead of the peak.
local_max <- vel |>
  dplyr::group_by(.data$sex) |>
  dplyr::filter(
    .data$velocity > dplyr::lag(.data$velocity),
    .data$velocity > dplyr::lead(.data$velocity),
    .data$time > 3
  ) |>
  dplyr::ungroup()
# Males also have a shallow mid-childhood maximum near 5 years; the
# pubertal peak is the larger one.
stopifnot(all(c("Female", "Male") %in% local_max$sex))
peaks <- local_max |>
  dplyr::group_by(.data$sex) |>
  dplyr::slice_max(.data$velocity, n = 1) |>
  dplyr::ungroup() |>
  dplyr::arrange(.data$sex) |>
  dplyr::mutate(published = c(8.5, 12.5)) |>
  dplyr::select("sex", "time", "velocity", "published")
stopifnot(all(abs(peaks$time - peaks$published) <= 0.5))
peaks |>
  dplyr::rename(
    "Sex" = "sex", "Model peak age (y)" = "time",
    "Peak velocity (cm/y)" = "velocity", "Published peak age (y)" = "published"
  ) |>
  knitr::kable(digits = 2)
```

```{r fig5, fig.cap = "Replicates Figure 5 of Hosey 2022: height velocity of the T2DM-specific height curves, females (left) and males (right). The published solid lines start near 9.3 (female) and 7.6 (male) cm/year at 2.5 years, peak near 6.6 cm/year at 8.5 years (female) and 8.9 cm/year at 12.5 years (male), and decline towards zero by 25 years."}
vel |>
  dplyr::filter(.data$time >= 2.5, .data$time <= 24.5) |>
  ggplot(aes(.data$time, .data$velocity)) +
  geom_line() +
  facet_wrap(~sex) +
  labs(x = "Age (years)", y = "Height velocity (cm/year)") +
  coord_cartesian(ylim = c(0, 10)) +
  theme_bw()
```

On the 0.01-year grid the model's pubertal peaks fall at 8.75 years (female,
6.66 cm/year) and 12.33 years (male, 9.0 cm/year). Both are within the 6-month
binning of the published 8.5 and 12.5 years. The start of the curves at 2.5
years (female 9.3, male 7.6 cm/year) also agrees with the Figure 5 panels. The
small female $b_2$ (0.0039 per year) is consistent with Figure 5 as printed.
It makes the second logistic almost linear, and it gives the slow residual
female velocity of about 0.1 cm/year still visible at 24 years.

## 3. Curves against Figure 2

The maintainers digitised the red "patient-specific model" curves of Figure 2 at
whole-year ages, from the article PDF rendered at 250 dpi with axes calibrated
on the tick marks. These values are read off a figure, not published numbers.
Reading error is about 0.5 cm for height and about 1 kg for weight.

```{r fig2-digitised}
fig2 <- tibble::tibble(
  time = rep(2:18, 4),
  sex = rep(c("Female", "Male", "Female", "Male"), each = 17),
  output = rep(c("ht", "ht", "bw", "bw"), each = 17),
  digitised = c(
    # Figure 2a, height, females
    80.6, 89.1, 97.9, 106.1, 113.5, 119.9, 126.1, 133.6, 140.4, 145.2, 149.3,
    153.9, 156.6, 158.4, 159.7, 160.7, 161.5,
    # Figure 2b, height, males
    83.5, 92.1, 98.7, 105.7, 113.4, 120.8, 127.2, 133.0, 138.6, 145.1, 151.9,
    158.9, 164.7, 168.5, 170.9, 172.6, 173.5,
    # Figure 2c, weight, females
    11.5, 13.7, 18.0, 21.7, 26.0, 31.2, 38.2, 44.9, 51.6, 57.9, 64.0, 70.7,
    78.3, 87.6, 96.1, 102.2, 104.1,
    # Figure 2d, weight, males
    13.2, 16.7, 20.4, 23.6, 26.0, 29.1, 34.9, 43.0, 51.4, 60.3, 69.4, 79.2,
    89.6, 98.5, 106.3, 116.3, 124.7
  )
)

cmp <- typ |>
  dplyr::filter(round(.data$time, 2) %in% 2:18) |>
  dplyr::mutate(time = round(.data$time)) |>
  dplyr::select("time", "sex", "ht", "bw") |>
  tidyr::pivot_longer(c("ht", "bw"), names_to = "output", values_to = "model") |>
  dplyr::inner_join(fig2, by = c("time", "sex", "output")) |>
  dplyr::mutate(pct_diff = 100 * (.data$model - .data$digitised) / .data$digitised)
stopifnot(nrow(cmp) == nrow(fig2))

# Height: the printed three-significant-figure logistic coefficients
# reproduce the figure at every age.
cmp_ht <- dplyr::filter(cmp, .data$output == "ht")
stopifnot(all(abs(cmp_ht$pct_diff) < 3))

# Weight: the S1 polynomial reproduces the figure over the core of the age
# range (4 to 16 years). The edges diverge; see the Errata.
cmp_bw_core <- dplyr::filter(cmp, .data$output == "bw", .data$time >= 4, .data$time <= 16)
stopifnot(
  abs(median(cmp_bw_core$pct_diff)) < 3,
  all(abs(cmp_bw_core$pct_diff) < 10)
)

cmp |>
  dplyr::filter(.data$time %in% c(2, 3, 4, 6, 8, 10, 12, 14, 16, 17, 18)) |>
  dplyr::mutate(output = ifelse(.data$output == "ht", "Height (cm)", "Weight (kg)")) |>
  dplyr::arrange(.data$output, .data$sex, .data$time) |>
  dplyr::rename(
    "Output" = "output", "Sex" = "sex", "Age (y)" = "time",
    "Model" = "model", "Figure 2 (digitised)" = "digitised", "% difference" = "pct_diff"
  ) |>
  knitr::kable(digits = 1)
```

```{r fig2, fig.height = 7, fig.cap = "Replicates Figure 2 of Hosey 2022: T2DM-specific height and weight curves (solid) with the model x (1 +/- CV) bands (dashed), and the maintainers' digitisation of the published red model curves (points). Panels are height and weight for females and males, as in Figure 2a-d."}
bands <- typ |>
  dplyr::filter(.data$time <= 18) |>
  dplyr::select("time", "sex", "ht", "bw", "propSd_ht", "propSd_bw") |>
  tidyr::pivot_longer(c("ht", "bw"), names_to = "output", values_to = "model") |>
  dplyr::mutate(
    cv = ifelse(.data$output == "ht", .data$propSd_ht, .data$propSd_bw),
    lo = .data$model * (1 - .data$cv),
    hi = .data$model * (1 + .data$cv)
  )
lab <- c(ht = "Height (cm)", bw = "Weight (kg)")
ggplot(bands, aes(.data$time)) +
  geom_line(aes(y = .data$model), colour = "firebrick") +
  geom_line(aes(y = .data$lo), colour = "firebrick", linetype = "dashed", alpha = 0.6) +
  geom_line(aes(y = .data$hi), colour = "firebrick", linetype = "dashed", alpha = 0.6) +
  geom_point(data = fig2, aes(y = .data$digitised), size = 1) +
  facet_grid(factor(output, levels = c("ht", "bw"), labels = lab) ~ sex, scales = "free_y") +
  labs(x = "Age (years)", y = NULL) +
  theme_bw()
```

The Figure 2 bands also confirm the residual-error form. The printed upper and
lower pink lines at 18 years sit at about 173 and 151 cm (female height, curve
162 cm) and at about 172 and 78 kg (male weight, curve about 125 kg). Those are
the curve multiplied by $1 \pm$ the Table 2 validation-set CVs (6.75% and
37.8%). This fixes both the proportional form and the Table 2 column the
authors used.

## 4. Residual variability of the stochastic cohort

The spread of `sim / pred` in the virtual cohort must recover the encoded CVs.
Both sides come from the same draw, so the only difference is Monte-Carlo
error. With 1,400 draws per sex and output, the standard error of an SD is
about 2% of its value, and a 10% tolerance leaves ample room.

```{r vpc-check}
cv_chk <- vpc |>
  dplyr::group_by(.data$output, .data$sex) |>
  dplyr::summarise(
    n = dplyr::n(),
    cv_sim = stats::sd(.data$sim / .data$pred),
    .groups = "drop"
  ) |>
  dplyr::mutate(cv_table2 = c(0.0675, 0.0587, 0.459, 0.378))
stopifnot(
  all(cv_chk$n == n_per_sex * length(ages_vpc)),
  all(abs(cv_chk$cv_sim / cv_chk$cv_table2 - 1) < 0.10)
)
cv_chk |>
  dplyr::rename(
    "Output" = "output", "Sex" = "sex", "N draws" = "n",
    "Simulated CV" = "cv_sim", "Table 2 CV" = "cv_table2"
  ) |>
  knitr::kable(digits = 3)
```

```{r vpc-plot, fig.height = 6, fig.cap = "Virtual cross-sectional cohort (200 per sex at each age) simulated with the proportional residual error. This is analogous to the Simcyp simulation of Supplemental Information S3. Lines are the typical curves."}
ggplot(vpc, aes(.data$time, .data$sim)) +
  geom_jitter(width = 0.2, alpha = 0.15, size = 0.6) +
  geom_line(
    data = bands |>
      dplyr::mutate(output = ifelse(.data$output == "ht", "Height (cm)", "Weight (kg)")),
    aes(.data$time, .data$model), colour = "firebrick"
  ) +
  facet_grid(output ~ sex, scales = "free_y") +
  labs(x = "Age (years)", y = NULL) +
  theme_bw()
```

Because the error is proportional and normal, a weight CV near 40-46% gives a
small fraction of non-physiological (near-zero or negative) weights in the lower
tail. The paper's Simcyp simulation carried the same CVs; users who need
strictly positive draws should truncate or resample them.

# Assumptions and deviations

* **Age is the time axis.** The rxode2 `time` variable is age in years. The
  model has no dosing and no ODE states. To use it as a covariate generator for
  a PBPK or popPK model, solve it on the age grid of interest and pass `ht` /
  `bw` on as `HT` / `WT` covariates.
* **Weight coefficients come from Supplemental Information S1.** The main-text
  weight equations are "truncated" to three significant figures and are not
  usable: they give about 420 kg for a 10-year-old boy. The paper itself says at
  least six significant figures are needed, and S1 supplies them.
* **The printed weight polynomials drift from Figure 2 at the edges of the age
  range.** Section 3 shows agreement within a few percent from 4 to 16 years.
  Outside that range the printed coefficients do not reproduce the published
  curve:
  - Males at 2 years: the polynomial gives 16.1 kg against about 13 kg in Figure 2d. It also dips slightly between 2 and 3 years.
  - 17 to 18 years: the polynomial turns down (male 96.8 kg and female 91.9 kg at 18 years) while Figure 2 keeps rising (about 125 and 104 kg).

  A tenth-degree polynomial amplifies rounding in its high-order coefficients.
  At 18 years the worst-case effect of rounding to six significant figures is
  about +/-42 kg for males, which can explain the male gap. For females it is
  about +/-7 kg, which cannot fully explain the gap. The coefficients are kept
  exactly as published and are **not** refitted to the figure. **Treat weights
  outside about 4 to 16 years with caution.** The female polynomial is not
  meaningful beyond about 22 years (it becomes negative before 25 years).
* **Height coefficients are the three-significant-figure values of the Results
  equations.** No higher-precision height coefficients were published. The
  logistic form is well conditioned, and these values reproduce Figure 2a-b to
  within about 1% and Figure 5 closely.
* **Residual error is proportional, with the validation-set CVs.** Table 2 gives
  CVs both for the validation set fitted to the final model and for the
  cross-validation folds (for example, female weight 45.9% versus 44.6 +/- 0.92%).
  The validation-set column is used because it reproduces the Figure 2 SD bands.
  The CV is a pooled statistic, RMSE divided by the mean observation (for
  example, 26.3 kg / 0.459 = 57 kg mean female weight). Encoding it as a
  constant proportional error is the authors' own use (Figure 2, Supplemental
  Information S3), not a per-age variance model.
* **Not extracted:** the CDC 50th-percentile comparator curves (not reproduced
  in the paper), the z-score analyses (Figures 3 and 4, descriptive), and the
  preliminary race/ethnicity curves of Supplemental Information S5 (no
  coefficients published).
* **Errata:** a check of EuropePMC and Crossref on 2026-09-30 found no erratum or correction for
  this article.
