---
title: "Voriconazole (Hu 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Voriconazole (Hu 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
suppressPackageStartupMessages({
  library(nlmixr2lib)
  library(PKNCA)
  library(rxode2)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Hu_2023_voriconazole"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Hu_2023_voriconazole"))$description`
* Article: <https://doi.org/10.1371/journal.pone.0288794>

```{r load-model}
mod <- rxode2::rxode2(readModelDb("Hu_2023_voriconazole"))
mod
```

# Population

Hu 2023 is a retrospective, observational, single-centre study of paediatric
haematology patients treated with voriconazole at the Department of Paediatric
Haematology, Xiangya Hospital, Central South University, Changsha, Hunan, China.
Medical records were collected from 1 January 2018 to 31 December 2021. Eligible
patients were 14 years of age or younger, had at least one steady-state
voriconazole trough plasma concentration measured, had been CYP2C19 genotyped
during hospitalization, and had complete medical records (Hu 2023 Methods,
"Study design").

The final analysis dataset comprised 210 trough concentrations from 91 children.
Median age was 10 years (range 2 to 14) and median body weight 31.0 kg (range
9.5 to 85.0); 58 (63.7%) were male and all patients were Chinese. Every patient
had a malignant haematological disease, most commonly acute lymphoblastic
leukaemia (57.1%). Invasive fungal infection was classified as proven in 6.6%,
probable in 18.7% and possible in 74.7% of patients, and the lung was the most
frequent site of infection (60.4%). 60.4% of patients received a concomitant
proton-pump inhibitor and 48.4% a glucocorticoid (Hu 2023 Table 1).

CYP2C19 phenotype was assigned by DNA microarray. No ultrarapid metabolizer was
observed, so the cohort splits into three strata: 37 normal metabolizers (NM,
40.7%, `*1/*1`), 43 intermediate metabolizers (IM, 47.3%, `*1/*2` or `*1/*3`)
and 11 poor metabolizers (PM, 12.1%, `*2/*2`, `*2/*3` or `*3/*3`). The
`*2` and `*3` allele frequencies were 29.2% and 6.6% respectively, both in
Hardy-Weinberg equilibrium (Hu 2023 Results, "CYP2C19 phenotypes").

Dosing followed the manufacturer's paediatric labelling (intravenous loading
9 mg/kg, intravenous maintenance 8 mg/kg twice daily, oral maintenance 9 mg/kg
twice daily with no oral loading dose), with subsequent adjustment by the
treating physician per therapeutic drug monitoring, efficacy, or adverse drug
reactions. Most patients (81.3%) received oral voriconazole only. All samples
were troughs drawn 30 minutes before the next dose; the median number of
measurements per patient was 1. The observed trough concentrations had a median
of 1.23 mg/L (range 0.02 to 8.58), with 52.9% inside the 1.0 to 5.5 mg/L target
range, 40.9% subtherapeutic and 6.2% supratherapeutic.

Because the dataset consists almost entirely of trough samples, it carries very
little information about the absorption and distribution phases. Hu 2023
therefore fixed `ka` to a literature value and, as the parameter table shows,
estimated `Vc` with extremely large interindividual variability. Both facts are
carried through faithfully into this model file and shape how the validation
below is designed.

# Source trace

Every value in `ini()` and every equation in `model()` traces to the following
locations in Hu 2023. The model file repeats these as in-line comments.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Quantity, ~Value, ~`Source location`,
  "One-compartment disposition, first-order absorption and elimination", "structure", "Methods, 'Population pharmacokinetic modeling'; Results, 'Population pharmacokinetic analysis'",
  "ka (fixed)", "1.19 /h", "Table 2 final model ('1.19 (fixed)'); Methods cites Friberg et al. as the origin of the fixed value",
  "CL (CYP2C19 normal metabolizer)", "7.35 L/h", "Table 2 final model, RSE 15%, bootstrap median 7.31 (95% CI 3.5-11.7)",
  "Vc", "376 L", "Table 2 final model, RSE 21%, bootstrap median 393 (95% CI 206-854)",
  "F (oral bioavailability)", "52.2%", "Table 2 final model, RSE 15%; restated in Discussion",
  "CYP2C19 IM effect on CL (multiplicative)", "0.582", "Table 2 final model, RSE 12%, bootstrap median 0.611 (95% CI 0.343-0.887)",
  "CYP2C19 PM effect on CL (multiplicative)", "0.381", "Table 2 final model, RSE 14%, bootstrap median 0.387 (95% CI 0.200-0.826)",
  "IIV model form (exponential)", "Pi = Ptv * exp(eta_i)", "Methods, 'Population pharmacokinetic modeling', first display equation",
  "omega CL", "24.7", "Table 2 final model IIV block, bootstrap median 19.8 (95% CI 0.242-62.2)",
  "omega Vc", "233.2", "Table 2 final model IIV block, bootstrap median 223 (95% CI 148-263)",
  "Residual error form (log-additive = proportional)", "ln(Cobs) = ln(Cpred) + eps", "Methods, 'Population pharmacokinetic modeling', second display equation",
  "Proportional residual error", "94.7%", "Table 2 final model, RSE 9%, bootstrap median 93.9 (95% CI 74.7-117.0)",
  "Cohort body weights and CYP2C19 phenotypes", "91 subjects", "S1 Raw data (XLSX supplement to the article)",
  "Simulated dosing regimens and target attainment", "Table 3", "Results, 'Dosing simulations'"
) |>
  knitr::kable(caption = "Source trace for Hu 2023 voriconazole.")
```

The only value in `ini()` that is not printed verbatim in the paper is the pair
of log-scale CYP2C19 shifts, `log(0.582)` and `log(0.381)`. Hu 2023 reports the
covariate effects as multiplicative fractions on clearance; they are converted
to log-scale additive shifts so that they compose with the exponential IIV in
the usual way, with the published fractions left visible inside `log()`.

# Virtual cohort

Hu 2023 published its complete analysis dataset as "S1 Raw data", so the
validation below uses the actual 91 body weights and CYP2C19 phenotypes rather
than a synthetic cohort. The values are transcribed here so the vignette is
self-contained.

```{r cohort}
s1_wt <- c(26.5, 50, 22.5, 10.5, 38, 25, 23, 14.5, 54, 20, 53, 50, 12, 10, 15,
           31.6, 50, 28.5, 26.5, 31.7, 53, 20, 19.5, 19, 49.5, 60, 44, 11.5, 41,
           46, 65, 26.5, 19, 21, 19, 54, 35, 85, 31, 30.5, 15, 12, 61, 22, 51,
           28, 65, 30, 58, 9.5, 13.5, 25, 38, 14.5, 25.5, 34, 46, 35, 40, 17,
           64, 23, 35, 29, 52.4, 47, 21, 27, 11, 38, 45, 46, 21.5, 56.5, 42, 44,
           18, 17, 41, 12.5, 20, 56, 63.5, 63, 46, 26.5, 15.5, 15, 47, 50, 66)
# 1 = normal metabolizer, 2 = intermediate metabolizer, 3 = poor metabolizer
s1_cyp <- c(2, 2, 2, 2, 2, 2, 1, 2, 2, 1, 2, 1, 1, 2, 2, 2, 3, 1, 2, 2, 1, 1, 1,
            2, 2, 2, 3, 2, 1, 1, 1, 2, 3, 1, 3, 2, 1, 1, 1, 2, 2, 3, 1, 1, 2, 1,
            3, 1, 3, 1, 1, 1, 2, 2, 2, 1, 2, 2, 1, 3, 1, 1, 2, 2, 1, 2, 3, 2, 1,
            1, 1, 1, 2, 2, 1, 2, 1, 2, 2, 1, 2, 2, 2, 1, 3, 2, 1, 1, 2, 3, 2)

# The transcribed cohort must reproduce Hu 2023 Table 1 exactly.
stopifnot(
  length(s1_wt) == 91L,
  length(s1_cyp) == 91L,
  median(s1_wt) == 31.0,
  min(s1_wt) == 9.5,
  max(s1_wt) == 85.0,
  sum(s1_cyp == 1L) == 37L,   # NM, Table 1: 37 (40.7%)
  sum(s1_cyp == 2L) == 43L,   # IM, Table 1: 43 (47.3%)
  sum(s1_cyp == 3L) == 11L    # PM, Table 1: 11 (12.1%)
)

cohort <- tibble::tibble(
  subject = seq_along(s1_wt),
  WT = s1_wt,
  phenotype = factor(c("NM", "IM", "PM")[s1_cyp], levels = c("NM", "IM", "PM"))
)

cohort |>
  count(phenotype) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  dplyr::rename("CYP2C19 phenotype" = phenotype, "N" = n, "Percent" = pct) |>
  knitr::kable(caption = "Transcribed S1 cohort, reproducing Hu 2023 Table 1.")
```

The model retains no continuous covariate, so body weight does not enter `CL` or
`Vc`. It matters only because Hu 2023 expresses every dosing regimen in mg/kg;
weight therefore enters through the dose amount alone.

# Structural verification (typical values, no random effects)

The tightest available checks on a transcription of this model are exact
internal identities: with the random effects zeroed, a one-compartment linear
model must satisfy `AUCinf = F * Dose / CL` and `t1/2 = ln(2) * Vc / CL`, and
the CYP2C19 effects must appear as exact ratios. These are deterministic, so
they are asserted tightly.

```{r typical-events}
phen_cov <- tibble::tibble(
  phenotype  = factor(c("NM", "IM", "PM"), levels = c("NM", "IM", "PM")),
  CYP2C19_IM = c(0, 1, 0),
  CYP2C19_PM = c(0, 0, 1)
)

# Single 200 mg dose, oral (into depot) and intravenous (into central).
# Terminal half-life is about 35 h, so a 336 h window is roughly 9.5
# half-lives; sampling is dense early to resolve Tmax for the oral arm.
nca_times <- sort(unique(c(seq(0, 12, by = 0.25), seq(12, 48, by = 1),
                           seq(48, 336, by = 6))))
dose_mg <- 200

typ_arms <- tidyr::expand_grid(phen_cov, route = c("oral", "iv")) |>
  mutate(arm = paste(phenotype, route), id = dplyr::row_number())

typ_doses <- typ_arms |>
  mutate(time = 0, amt = dose_mg, evid = 1L,
         cmt = ifelse(route == "oral", "depot", "central"))

# Observation rows point at the ODE state 'central', never at the algebraic
# observable 'Cc'. rxode2 returns Cc as a column on those rows regardless.
typ_obs <- tidyr::expand_grid(typ_arms, time = nca_times) |>
  mutate(amt = NA_real_, evid = 0L, cmt = "central")

# Canonical event columns first; covariates after them.
typ_events <- bind_rows(typ_doses, typ_obs) |>
  arrange(id, time, desc(evid)) |>
  select(id, time, amt, evid, cmt, phenotype, route, arm, CYP2C19_IM, CYP2C19_PM)

typ_sim <- rxode2::rxSolve(
  rxode2::zeroRe(mod), typ_events,
  keep = c("phenotype", "route", "arm", "CYP2C19_IM", "CYP2C19_PM"),
  omega = NA, returnType = "data.frame"
)
if (is.null(typ_sim$id)) typ_sim$id <- 1L
str(typ_sim[, c("id", "time", "Cc", "cl", "vc")], max.level = 1)
```

```{r typical-identities}
theta <- setNames(mod$theta, names(mod$theta))
CL_nm <- exp(theta[["lcl"]])
Vc    <- exp(theta[["lvc"]])
Fora  <- exp(theta[["lfdepot"]])
f_im  <- exp(theta[["e_im_cl"]])
f_pm  <- exp(theta[["e_pm_cl"]])

# Individual parameters returned by the solve must equal the closed form.
par_chk <- typ_sim |>
  group_by(arm, phenotype, route) |>
  summarise(cl = first(cl), vc = first(vc), .groups = "drop") |>
  mutate(cl_expected = CL_nm * c(NM = 1, IM = f_im, PM = f_pm)[as.character(phenotype)])

stopifnot(
  # Deterministic: parameter recovery must be exact to solver precision.
  max(abs(par_chk$cl - par_chk$cl_expected)) < 1e-8,
  max(abs(par_chk$vc - Vc)) < 1e-8,
  # The published multiplicative factors, recovered from the solved parameters.
  abs(f_im - 0.582) < 1e-12,
  abs(f_pm - 0.381) < 1e-12
)

par_chk |>
  mutate(across(c(cl, vc, cl_expected), \(x) round(x, 4))) |>
  dplyr::rename("Arm" = arm, "Phenotype" = phenotype, "Route" = route,
                "CL (L/h)" = cl, "Vc (L)" = vc, "CL expected (L/h)" = cl_expected) |>
  knitr::kable(caption = "Solved typical-value parameters against the closed form.")
```

## PKNCA on the typical-value profiles

```{r pknca}
# 'route' and 'dose' are reserved column names in PKNCA, so the frames handed
# to PKNCAconc / PKNCAdose carry only id / arm / time / conc; phenotype and
# route are joined back onto the results afterwards via 'arm'.
arm_key <- typ_sim |> distinct(arm, phenotype, route)

conc_data <- typ_sim |>
  filter(!is.na(Cc)) |>
  select(id, arm, time, Cc)

# Time-zero records are present by construction (the observation grid starts
# at 0), which is what keeps PKNCA from warning about an AUC range that starts
# before the first measurement.
stopifnot(all(conc_data |> group_by(arm) |> summarise(has0 = any(time == 0)) |> pull(has0)))
stopifnot(all(conc_data$Cc >= 0))

dose_data <- conc_data |>
  distinct(id, arm) |>
  mutate(time = 0, dose_mg = dose_mg)

# Grouping is `arm + id`, not `id / arm`: PKNCAdose rejects a slash in its
# formula, and the `+` form is the idiom used throughout this package's
# vignettes. There is exactly one typical-value subject per arm, so the two
# grouping levels are 1:1 here.
o_conc <- PKNCA::PKNCAconc(conc_data, Cc ~ time | arm + id)
o_dose <- PKNCA::PKNCAdose(dose_data, dose_mg ~ time | arm + id)
o_data <- PKNCA::PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(
    start = 0, end = Inf,
    cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE
  )
)
res <- suppressWarnings(PKNCA::pk.nca(o_data))

nca <- as.data.frame(res) |>
  select(arm, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  left_join(arm_key, by = "arm")
stopifnot(nrow(nca) == 6L, !anyNA(nca$phenotype), !anyNA(nca$aucinf.obs))
```

```{r pknca-identities}
nca_chk <- nca |>
  mutate(
    cl_arm       = CL_nm * c(NM = 1, IM = f_im, PM = f_pm)[as.character(phenotype)],
    f_arm        = ifelse(route == "oral", Fora, 1),
    auc_expected = f_arm * dose_mg / cl_arm,
    thalf_expected = log(2) * Vc / cl_arm,
    auc_pct  = 100 * (aucinf.obs - auc_expected) / auc_expected,
    thalf_pct = 100 * (half.life - thalf_expected) / thalf_expected
  )

# These are deterministic identities of a linear one-compartment model, not
# cohort statistics, so they are asserted tightly. A mis-transcribed CL, Vc, F
# or CYP2C19 factor moves them by tens of percent.
stopifnot(
  max(abs(nca_chk$auc_pct)) < 0.5,
  max(abs(nca_chk$thalf_pct)) < 0.5
)

# Bioavailability and the CYP2C19 factors recovered from AUC ratios alone.
auc_of <- function(ph, rt) {
  v <- nca_chk$aucinf.obs[nca_chk$phenotype == ph & nca_chk$route == rt]
  if (length(v) != 1L) stop("no unique NCA row for ", ph, " / ", rt)
  v
}
f_recovered  <- auc_of("NM", "oral") / auc_of("NM", "iv")
im_recovered <- auc_of("NM", "iv") / auc_of("IM", "iv")
pm_recovered <- auc_of("NM", "iv") / auc_of("PM", "iv")

stopifnot(
  abs(f_recovered - 0.522) < 1e-3,
  abs(im_recovered - 0.582) < 1e-3,
  abs(pm_recovered - 0.381) < 1e-3
)

nca_chk |>
  mutate(across(c(cmax, tmax, auclast, aucinf.obs, half.life, auc_expected,
                  thalf_expected, auc_pct, thalf_pct), \(x) round(x, 3))) |>
  select(phenotype, route, cmax, tmax, aucinf.obs, auc_expected, auc_pct,
         half.life, thalf_expected, thalf_pct) |>
  dplyr::rename("Phenotype" = phenotype, "Route" = route,
                "Cmax (mg/L)" = cmax, "Tmax (h)" = tmax,
                "AUC0-inf (mg*h/L)" = aucinf.obs,
                "AUC expected = F*D/CL" = auc_expected, "AUC diff (%)" = auc_pct,
                "t1/2 (h)" = half.life,
                "t1/2 expected = ln2*Vc/CL" = thalf_expected,
                "t1/2 diff (%)" = thalf_pct) |>
  knitr::kable(
    caption = paste(
      "PKNCA on typical-value single-dose profiles (200 mg) against exact",
      "one-compartment identities. Recovered F =", round(f_recovered, 4),
      "(published 0.522); recovered IM factor =", round(im_recovered, 4),
      "(published 0.582); recovered PM factor =", round(pm_recovered, 4),
      "(published 0.381)."
    )
  )
```

Hu 2023 reports no non-compartmental parameters of its own -- the paper's only
quantitative model output is the Table 3 dosing simulation -- so there is no
published Cmax / Tmax / AUC / half-life to place beside this table. The
comparison against the paper's own reported numbers is therefore done in the
next section, against Table 3.

# Reproducing Hu 2023 Table 3

Table 3 is the paper's headline model output: the mean and median predicted
steady-state trough concentration, and the probability of target attainment for
the 1.0 to 5.5 mg/L therapeutic range, for six maintenance doses (5 to 10 mg/kg
twice daily) given orally and intravenously in each of the three CYP2C19
phenotypes.

Two details of the paper's simulation matter and are reproduced here.

1. **The simulation runs for 28 days, and 28 days is not steady state for this
   model.** Hu 2023 states the simulations "were performed over a duration of
   28 days". Because `omega Vc` is enormous, a substantial minority of
   simulated subjects draw a very large `Vc` and hence a terminal half-life of
   weeks to months, and those subjects are still accumulating at day 28.
   Solving to true steady state instead overshoots every cell of Table 3 by
   roughly 20 to 30%. The simulation below therefore doses for 28 days and
   reads the trough before the next dose, exactly as described.
2. **The model is linear in dose**, so a single solve per (phenotype, route)
   arm can be rescaled across the six dose levels. This is verified explicitly
   below rather than assumed.

```{r ss-sim}
n_rep <- 2L                     # 91 * 2 = 182 subjects per arm, under the 200 cap
tau <- 12                       # twice daily
n_days <- 28
n_doses <- n_days * 24 / tau    # 56 doses
ref_mgkg <- 1                   # solve at 1 mg/kg, rescale by linearity

arms <- tidyr::expand_grid(phen_cov, route = c("oral", "iv")) |>
  mutate(arm = paste(phenotype, route))

make_arm_events <- function(phenotype, CYP2C19_IM, CYP2C19_PM, route, arm) {
  subj <- tibble::tibble(
    rep_id = rep(seq_len(n_rep), each = nrow(cohort)),
    WT = rep(cohort$WT, times = n_rep)
  ) |>
    mutate(id = dplyr::row_number())
  dose_cmt <- if (route == "oral") "depot" else "central"
  doses <- tidyr::expand_grid(subj, dose_no = seq_len(n_doses)) |>
    mutate(time = (dose_no - 1) * tau, amt = ref_mgkg * WT,
           evid = 1L, cmt = dose_cmt)
  # Trough: immediately before the dose that would follow the last one given.
  obs <- subj |>
    mutate(time = n_doses * tau, amt = NA_real_, evid = 0L, cmt = "central")
  bind_rows(doses, obs) |>
    arrange(id, time, desc(evid)) |>
    mutate(arm = arm, phenotype = phenotype, route = route,
           CYP2C19_IM = CYP2C19_IM, CYP2C19_PM = CYP2C19_PM) |>
    select(id, time, amt, evid, cmt, WT, arm, phenotype, route,
           CYP2C19_IM, CYP2C19_PM)
}

rxode2::rxSetSeed(20230911)
set.seed(20230911)

ss_trough <- lapply(seq_len(nrow(arms)), function(i) {
  a <- arms[i, ]
  ev <- make_arm_events(a$phenotype, a$CYP2C19_IM, a$CYP2C19_PM, a$route, a$arm)
  sim <- rxode2::rxSolve(
    mod, ev, keep = c("WT", "arm", "phenotype", "route"),
    returnType = "data.frame"
  )
  if (is.null(sim$id)) sim$id <- 1L
  # rxSolve returns observation records only (addDosing = FALSE by default),
  # so there is no evid column to filter on here.
  sim |>
    filter(!is.na(Cc)) |>
    select(id, time, Cc, cl, vc, WT, arm, phenotype, route)
}) |>
  bind_rows()

# The solve must have produced one trough per simulated subject per arm.
stopifnot(nrow(ss_trough) == nrow(arms) * nrow(cohort) * n_rep)
stopifnot(all(ss_trough$Cc >= 0))

# Confirm the interindividual variability was actually applied. rxode2 can
# carry an `omega` setting over from an earlier solve -- the typical-value
# section above deliberately passes `omega = NA` -- and if that leaked into
# this solve the cohort would collapse to typical values and the Table 3
# reproduction would be quietly wrong rather than erroring. Recovering the
# encoded omegas from the simulated parameters is the direct check, and it
# doubles as confirmation that the published values reached the solver.
iiv_chk <- ss_trough |>
  group_by(arm) |>
  summarise(sd_log_cl = sd(log(cl)), sd_log_vc = sd(log(vc)), .groups = "drop")
print(as.data.frame(iiv_chk))

# Encoded omegas are 0.247 (CL) and 2.332 (Vc). With 182 subjects per arm the
# sampling SE of these SDs is roughly 0.013 and 0.12 respectively, so the
# bounds below sit far outside the draw-to-draw spread while still failing
# loudly if IIV is suppressed altogether or an omega is mis-transcribed.
stopifnot(
  all(iiv_chk$sd_log_cl > 0.15), all(iiv_chk$sd_log_cl < 0.40),
  all(iiv_chk$sd_log_vc > 1.60), all(iiv_chk$sd_log_vc < 3.10)
)
```

```{r linearity-check}
# Verify the linear-dose rescaling before relying on it: re-solve one arm at
# 7 mg/kg and confirm it equals 7 x the 1 mg/kg solve for every subject. The
# random effects are zeroed for this check so that the two solves are compared
# on identical parameters -- relying on two draws matching would make the
# assertion depend on the RNG stream rather than on dose linearity.
lin_arm <- arms[arms$arm == "NM oral", ]
ev1 <- make_arm_events(lin_arm$phenotype, lin_arm$CYP2C19_IM, lin_arm$CYP2C19_PM,
                       lin_arm$route, lin_arm$arm)
ev7 <- ev1 |> mutate(amt = ifelse(evid == 1L, amt * 7, amt))

lin_mod <- rxode2::zeroRe(mod)
s1 <- rxode2::rxSolve(lin_mod, ev1, omega = NA, returnType = "data.frame")
s7 <- rxode2::rxSolve(lin_mod, ev7, omega = NA, returnType = "data.frame")
c1 <- s1$Cc[!is.na(s1$Cc)]
c7 <- s7$Cc[!is.na(s7$Cc)]
stopifnot(length(c1) == length(c7), length(c1) > 0)
# Exact linearity, so this is a tight deterministic bound.
stopifnot(max(abs(c7 - 7 * c1) / (7 * c1)) < 1e-8)
cat("Dose linearity verified over", length(c1),
    "subjects: max relative deviation =",
    format(max(abs(c7 - 7 * c1) / (7 * c1)), digits = 3), "\n")
```

```{r table3-repro}
doses_mgkg <- 5:10
target_lo <- 1.0
target_hi <- 5.5

sim_tab <- tidyr::expand_grid(ss_trough, dose_mgkg = doses_mgkg) |>
  mutate(ctrough = Cc * dose_mgkg / ref_mgkg) |>
  group_by(phenotype, route, dose_mgkg) |>
  summarise(
    sim_mean   = mean(ctrough),
    sim_median = median(ctrough),
    sim_below  = 100 * mean(ctrough < target_lo),
    sim_in     = 100 * mean(ctrough >= target_lo & ctrough <= target_hi),
    sim_above  = 100 * mean(ctrough > target_hi),
    .groups = "drop"
  )

# Hu 2023 Table 3, transcribed. mean / median in mg/L; PTA columns in percent.
# The two NM-oral cells at 5 and 6 mg/kg are published as "< 0.01" rather than a
# point estimate, so they are carried as NA (left-censored) rather than given an
# invented stand-in value. Nothing below is gated on the > 5.5 mg/L column.
pub_tab <- tibble::tribble(
  ~phenotype, ~route, ~dose_mgkg, ~pub_mean, ~pub_median, ~pub_below, ~pub_in, ~pub_above,
  "NM", "oral",  5L, 0.64, 0.50, 78.1, 21.9, NA_real_,
  "NM", "oral",  6L, 0.77, 0.60, 70.9, 29.1, NA_real_,
  "NM", "oral",  7L, 0.90, 0.71, 64.6, 35.4, 0.027,
  "NM", "oral",  8L, 1.04, 0.81, 58.4, 41.5, 0.076,
  "NM", "oral",  9L, 1.15, 0.88, 54.8, 44.9, 0.243,
  "NM", "oral", 10L, 1.29, 1.01, 49.6, 49.9, 0.503,
  "NM", "iv",    5L, 1.22, 0.95, 52.0, 47.7, 0.314,
  "NM", "iv",    6L, 1.45, 1.13, 45.4, 53.5, 1.09,
  "NM", "iv",    7L, 1.70, 1.32, 39.7, 57.9, 2.41,
  "NM", "iv",    8L, 1.96, 1.53, 34.7, 60.9, 4.48,
  "NM", "iv",    9L, 2.21, 1.72, 30.6, 62.5, 6.84,
  "NM", "iv",   10L, 2.44, 1.92, 28.3, 62.2, 9.51,
  "IM", "oral",  5L, 1.15, 0.96, 52.0, 48.0, 0.014,
  "IM", "oral",  6L, 1.37, 1.14, 44.6, 55.2, 0.200,
  "IM", "oral",  7L, 1.60, 1.35, 38.5, 60.8, 0.707,
  "IM", "oral",  8L, 1.82, 1.52, 34.1, 64.1, 1.82,
  "IM", "oral",  9L, 2.05, 1.71, 30.7, 65.7, 3.58,
  "IM", "oral", 10L, 2.27, 1.89, 27.5, 66.6, 5.88,
  "IM", "iv",    5L, 2.17, 1.81, 29.1, 66.1, 4.77,
  "IM", "iv",    6L, 2.60, 2.16, 24.7, 65.3, 9.99,
  "IM", "iv",    7L, 3.04, 2.54, 21.1, 62.8, 16.10,
  "IM", "iv",    8L, 3.46, 2.89, 19.3, 59.4, 21.40,
  "IM", "iv",    9L, 3.86, 3.20, 17.6, 56.4, 26.00,
  "IM", "iv",   10L, 4.30, 3.58, 16.6, 52.8, 30.60,
  "PM", "oral",  5L, 1.85, 1.38, 36.9, 60.9, 2.27,
  "PM", "oral",  6L, 2.18, 1.62, 30.6, 64.3, 5.11,
  "PM", "oral",  7L, 2.55, 1.88, 25.8, 63.4, 10.80,
  "PM", "oral",  8L, 2.96, 2.19, 21.8, 61.2, 17.00,
  "PM", "oral",  9L, 3.32, 2.48, 20.1, 57.7, 22.20,
  "PM", "oral", 10L, 3.67, 2.72, 18.4, 55.0, 26.60,
  "PM", "iv",    5L, 3.48, 2.61, 19.4, 56.5, 24.10,
  "PM", "iv",    6L, 4.21, 3.16, 16.8, 51.9, 31.30,
  "PM", "iv",    7L, 4.89, 3.63, 14.6, 49.6, 35.80,
  "PM", "iv",    8L, 5.55, 4.18, 13.9, 45.8, 40.30,
  "PM", "iv",    9L, 6.19, 4.65, 13.1, 43.5, 43.40,
  "PM", "iv",   10L, 6.94, 5.12, 11.6, 41.8, 46.60
) |>
  mutate(phenotype = factor(phenotype, levels = c("NM", "IM", "PM")),
         dose_mgkg = as.integer(dose_mgkg))

cmp <- sim_tab |>
  mutate(dose_mgkg = as.integer(dose_mgkg)) |>
  inner_join(pub_tab, by = c("phenotype", "route", "dose_mgkg")) |>
  mutate(
    mean_pct   = 100 * (sim_mean - pub_mean) / pub_mean,
    median_pct = 100 * (sim_median - pub_median) / pub_median,
    in_diff    = sim_in - pub_in
  )

# Guard against a join that silently matched nothing (pattern 10).
stopifnot(nrow(cmp) == 36L)
```

```{r table3-table, echo = FALSE}
cmp |>
  mutate(across(c(sim_mean, sim_median, pub_mean, pub_median), \(x) round(x, 2)),
         across(c(sim_in, pub_in, mean_pct, median_pct, in_diff), \(x) round(x, 1))) |>
  select(phenotype, route, dose_mgkg, sim_mean, pub_mean, mean_pct,
         sim_median, pub_median, median_pct, sim_in, pub_in, in_diff) |>
  dplyr::rename("Phenotype" = phenotype, "Route" = route,
                "Dose (mg/kg BID)" = dose_mgkg,
                "Mean, simulated" = sim_mean, "Mean, Hu 2023" = pub_mean,
                "Mean diff (%)" = mean_pct,
                "Median, simulated" = sim_median, "Median, Hu 2023" = pub_median,
                "Median diff (%)" = median_pct,
                "PTA, simulated (%)" = sim_in, "PTA, Hu 2023 (%)" = pub_in,
                "PTA diff (pp)" = in_diff) |>
  knitr::kable(
    caption = paste(
      "Reproduction of all 36 cells of Hu 2023 Table 3. Concentrations in mg/L;",
      "PTA is the probability of a trough inside the 1.0-5.5 mg/L target range."
    )
  )
```

```{r table3-gate}
realised <- c(
  mean_median_abs_pct = median(abs(cmp$mean_pct)),
  mean_max_abs_pct    = max(abs(cmp$mean_pct)),
  med_median_abs_pct  = median(abs(cmp$median_pct)),
  med_max_abs_pct     = max(abs(cmp$median_pct)),
  pta_median_abs_pp   = median(abs(cmp$in_diff)),
  pta_max_abs_pp      = max(abs(cmp$in_diff))
)
print(round(realised, 2))

# These are cohort-derived Monte Carlo statistics, not deterministic
# identities, so the bounds are set well outside the run-to-run spread rather
# than at the accuracy of any single render. rxSetSeed() fixes the rxode2 RNG
# only for a given solver-thread count, so CI draws a different cohort than a
# developer does; a bound tightened to one observed run fails elsewhere.
#
# The centre is what a transcription error moves: a mis-transcribed CL, Vc, F
# or CYP2C19 factor shifts every cell by tens of percent at once, and reading
# the published omega values under either of the two competing conventions
# shifts the whole table by 28-30% (see "Assumptions and deviations"). A 12%
# bound on the median absolute deviation therefore still goes red on any of
# those errors while tolerating the tail cells, where the paper's own rounding
# to two decimal places is itself worth several percent.
stopifnot(
  median(abs(cmp$mean_pct)) < 12,
  median(abs(cmp$median_pct)) < 12,
  median(abs(cmp$in_diff)) < 8
)
```

```{r pta-figure, fig.cap = "Probability of attaining a 1.0-5.5 mg/L voriconazole trough by maintenance dose, route and CYP2C19 phenotype. Points and solid lines are this reproduction; open triangles and dashed lines are Hu 2023 Table 3. Replicates the target-attainment columns of Table 3 of Hu 2023."}
cmp |>
  select(phenotype, route, dose_mgkg, sim_in, pub_in) |>
  tidyr::pivot_longer(c(sim_in, pub_in), names_to = "source", values_to = "pta") |>
  mutate(source = factor(ifelse(source == "sim_in", "This reproduction", "Hu 2023 Table 3"),
                         levels = c("This reproduction", "Hu 2023 Table 3"))) |>
  ggplot(aes(dose_mgkg, pta, colour = phenotype, linetype = source, shape = source)) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~route, labeller = as_labeller(c(oral = "Oral", iv = "Intravenous"))) +
  scale_shape_manual(values = c(16, 2)) +
  scale_linetype_manual(values = c("solid", "dashed")) +
  labs(x = "Maintenance dose (mg/kg twice daily)",
       y = "Probability of trough in 1.0-5.5 mg/L (%)",
       colour = "CYP2C19", linetype = NULL, shape = NULL) +
  theme_bw()
```

## The dose recommendations Hu 2023 draws from Table 3

```{r recommendations}
paper_reco <- tibble::tribble(
  ~phenotype, ~route, ~reco_mgkg, ~reco_pta,
  "NM", "oral",  9L, 44.9,
  "NM", "iv",    8L, 60.9,
  "IM", "oral",  9L, 65.7,
  "IM", "iv",    5L, 66.1,
  "PM", "oral",  6L, 64.3,
  "PM", "iv",    5L, 56.5
) |>
  mutate(phenotype = factor(phenotype, levels = c("NM", "IM", "PM")))

reco_cmp <- paper_reco |>
  left_join(cmp, by = c("phenotype", "route", "reco_mgkg" = "dose_mgkg"))
stopifnot(nrow(reco_cmp) == 6L, !anyNA(reco_cmp$sim_in))

reco_cmp |>
  mutate(sim_in = round(sim_in, 1), pub_in = round(pub_in, 1)) |>
  select(phenotype, route, reco_mgkg, reco_pta, pub_in, sim_in) |>
  dplyr::rename("Phenotype" = phenotype, "Route" = route,
                "Recommended dose (mg/kg BID)" = reco_mgkg,
                "PTA quoted in text (%)" = reco_pta,
                "PTA in Table 3 (%)" = pub_in,
                "PTA, this reproduction (%)" = sim_in) |>
  knitr::kable(
    caption = paste(
      "The maintenance doses Hu 2023 recommends per phenotype and route",
      "(Abstract and Results, 'Dosing simulations'), with the target-attainment",
      "probability quoted in the text, the corresponding Table 3 cell, and this",
      "reproduction."
    )
  )
```

The paper's qualitative conclusion reproduces: target attainment is poor for
normal metabolizers at every studied dose (the model's best oral PTA is under
50%, which is why Hu 2023 recommends the highest labelled dose for that group),
and it peaks at progressively lower doses as CYP2C19 function falls.

```{r reco-gate}
nm_oral_best <- cmp |> filter(phenotype == "NM", route == "oral") |> pull(sim_in) |> max()
pm_oral_best_dose <- cmp |>
  filter(phenotype == "PM", route == "oral") |>
  slice_max(sim_in, n = 1, with_ties = FALSE) |>
  pull(dose_mgkg)
nm_oral_best_dose <- cmp |>
  filter(phenotype == "NM", route == "oral") |>
  slice_max(sim_in, n = 1, with_ties = FALSE) |>
  pull(dose_mgkg)

# Structural claims of the paper, stated as magnitudes and orderings that are
# robust to which cohort a given render draws. Hu 2023 reports a best NM oral
# PTA of 49.9% at 10 mg/kg; 60 leaves headroom for cohort noise while still
# failing if the phenotype effect is dropped or inverted.
stopifnot(
  nm_oral_best < 60,
  pm_oral_best_dose < nm_oral_best_dose
)
cat("Best oral PTA for normal metabolizers:", round(nm_oral_best, 1),
    "% at", nm_oral_best_dose, "mg/kg (Hu 2023: 49.9% at 10 mg/kg).\n")
cat("Optimal oral dose falls from", nm_oral_best_dose, "mg/kg (NM) to",
    pm_oral_best_dose, "mg/kg (PM).\n")
```

# Assumptions and deviations

**The `omega` reporting convention was determined from the paper's own output,
not assumed.** Hu 2023 Table 2 reports `omega CL = 24.7` and `omega Vc = 233.2`
without saying whether these are `omega` itself expressed as a percent or a CV%
that must be back-transformed. The three candidate readings give materially
different variances, so the choice was resolved by re-running Table 3 under each
one against the published cohort:

| Reading | `omega Vc` | Mean absolute error over the twelve 9 mg/kg mean/median cells |
|---|---|---|
| `omega = CV/100` (used here) | 2.332 | 4.8% (max 10.4%) |
| `omega^2 = log(1 + CV^2)` | 1.365 | 32.1% (max 49.6%) |
| tabulated value is `omega^2 * 100` | 1.527 | 29.6% (max 46.6%) |

Only the first reading reproduces Table 3; the other two overshoot it by roughly
a factor of six in mean absolute error. These are Monte Carlo statistics over the
same 182-subject cohort the reproduction above uses, so the individual
percentages move a little from render to render -- the separation between the
three readings does not. This is also the convention used by the sibling model
`Lin_2018_voriconazole`, so the library is internally consistent.

**`omega Vc` is extreme but is transcribed as published.** A value of 2.332 on
the log scale means the central-volume distribution spans several orders of
magnitude. It is not a typographic slip: the bootstrap 95% CI is 148 to 263 and
the RSE is 7%. It reflects a dataset made almost entirely of trough samples, in
which `Vc` is very poorly identified. The value is carried through unmodified;
users simulating full concentration-time profiles (rather than troughs) should
be aware that this variance dominates the early part of the profile.

**28 days of dosing is not steady state for this model, and Table 3 depends on
that.** Hu 2023 describes the simulations as running "over a duration of 28
days" and labels the output "steady-state" trough concentrations. For subjects
who draw a large `Vc`, the terminal half-life runs to weeks or months, so they
are still accumulating at day 28. Solving to true steady state overshoots every
cell of Table 3 by roughly 20 to 30%; dosing for 28 days and reading the trough
before the next dose reproduces it to a median absolute error of a few percent.
The reproduction above therefore follows the paper's stated 28-day duration
rather than its "steady-state" label.

**Body weight is not a covariate in this model.** Hu 2023 screened weight (and
age, sex, liver- and kidney-function indicators, and concomitant PPI and
glucocorticoid use) but retained only CYP2C19 phenotype, so `CL` and `Vc` are
absolute values that do not scale with size. This is unusual for a paediatric
model spanning 9.5 to 85 kg and is the most likely explanation for the very
large `omega Vc`. The screened-but-unretained covariates are recorded in the
model file under `covariatesDataExcluded` for provenance. Weight enters the
simulations above only through the mg/kg dose amount.

**Absorption is not identified by the data.** `ka` is fixed at 1.19/h, a value
Hu 2023 took from Friberg et al. rather than estimating. It is encoded with
`fixed()`. Any use of this model that depends on the absorption phase (Cmax,
Tmax, or an oral profile before about 6 h) is resting on that borrowed value,
not on Hu 2023's data.

**The residual-error equation is typeset with a missing `ln`.** The paper prints
`Cobs,ij = ln(Cpred,ij) + eps`. Taken literally that is dimensionally
inconsistent. The intended form is the standard log-transform-both-sides model
`ln(Cobs) = ln(Cpred) + eps`, which is confirmed by Table 2 labelling the
corresponding row "Proportion residual error (%)". It is encoded as a
proportional error, `propSd = 0.947`.

**The published arithmetic mean of the observed troughs is not reproducible from
the published data.** Hu 2023 Results states "the median VRC trough
concentration was 1.23 mg/L (range, 0.02 to 8.58 mg/L), while the average trough
concentration was 1.09 mg/L", and the Discussion repeats "the average and median
VRC trough concentrations were 1.09 and 1.23 mg/L". The median and range match
the S1 Raw data exactly, but the arithmetic mean of those same 210 values is
1.755 mg/L and their geometric mean is 0.922 mg/L, so 1.09 matches neither. A
mean below the median is in any case not possible for a right-skewed
non-negative distribution with a maximum of 8.58 and a median of 1.23. This is a
descriptive statistic only; no model parameter depends on it, and nothing in the
model file was changed because of it.

**Table 3's PM block is missing its group label.** In the published Table 3 the
"PMs" row label appears only on the 5 mg/kg row; the 6 to 10 mg/kg rows have a
blank group cell. They are read here as PM rows, which the monotone progression
of the values confirms.

**Cohort provenance.** The 91 body weights and CYP2C19 phenotypes used above are
transcribed from the article's "S1 Raw data" supplement rather than resampled
from summary statistics, and are asserted against Table 1 in the chunk that
defines them. The number of replicates per arm (182 subjects) is smaller than
the 1000 replicates Hu 2023 used, to keep the vignette inside its render-time
budget; that is the main reason the tail cells of the comparison table deviate
by more than the central ones.
