---
title: "Nivolumab and ipilimumab pediatric population PK (Hu 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nivolumab and ipilimumab pediatric population PK (Hu 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Hu 2024 developed two independent pooled adult-plus-pediatric population PK
models -- one for nivolumab, one for ipilimumab -- to answer a single question:
after body size has been accounted for, is there anything left of the
pediatric-versus-adult difference in monoclonal-antibody PK? The answer the
paper reaches is yes, and that the residual effect is tumour-type dependent.
Because the authors fitted the two drugs separately, on different datasets and
with different structural forms (nivolumab time-varying clearance, ipilimumab
stationary clearance), the extraction is two model files sharing this one
vignette.

```{r models}
nivo <- rxode2::rxode(readModelDb("Hu_2024_nivolumab"))
ipi  <- rxode2::rxode(readModelDb("Hu_2024_ipilimumab"))
```

* Citation: `r nivo$reference`
* Nivolumab model: `r nivo$description`
* Ipilimumab model: `r ipi$description`
* Article: <https://doi.org/10.1002/psp4.13098>
* Supplement (Appendix S1 supplementary results and tables, Appendix File S2
  and S3 NONMEM control streams):
  <https://www.ebi.ac.uk/europepmc/webservices/rest/PMC10941504/supplementaryFiles>

# Population

**Nivolumab** (Table 1). 13,104 concentrations from 2,325 patients across 13
studies, of whom 275 were pediatric. The analysis populations are adult
melanoma (n = 993, the reference), adult lymphoma (n = 274), adult
glioblastoma (n = 556), adult other solid tumours (n = 227), pediatric solid
tumours (n = 79), pediatric lymphoma (n = 46), and pediatric CNS tumours
(n = 150). Overall age median 55 years (range 1-90), baseline body weight
median 76.2 kg (range 9.3-168), 36.5% female, 90.3% White. The pediatric
cohorts are markedly smaller in both weight and lean body mass than the adults:
pediatric CNS-tumour median weight 33 kg with mean lean body mass 30.2 kg,
against an adult-melanoma mean weight of 82.3 kg and mean lean body mass of
57.7 kg.

**Ipilimumab** (Table 2). 6,020 concentrations from 1,427 patients across 10
studies, of whom 138 were pediatric: adult melanoma (n = 1261, the reference),
adult CNS tumours (n = 6), adult other solid tumours (n = 22), pediatric solid
tumours (n = 43), pediatric CNS tumours (n = 72), and pediatric melanoma
(n = 23). Overall age median 58 years (range 1-89), weight median 78.1 kg
(range 10.2-160), 36.9% female, 94.8% White.

Both analyses estimated lean body mass by the Boer equation in adults and the
Peter equation in children (Table 1 Note).

The same information is available programmatically from each model's
`population` metadata:

```{r population-metadata}
str(nivo$population, max.level = 1, give.attr = FALSE)
```

# Source trace

Every `ini()` entry in both model files carries an in-file comment naming its
source location. The table below collects them.

| Equation / parameter | Value | Source location |
|---|---|---|
| `CL_i(t) = CL0TV_i * exp(Emax_i * t^HILL / (T50^HILL + t^HILL)) * exp(eta_CL)` | n/a | Hu 2024 Results, first nivolumab equation |
| `CL0TV_i = CL0REF * (WTB/75)^CLWTB * (eGFR/90)^CLeGFR * exp(...)` | n/a | Hu 2024 Results, nivolumab `CL0TV,i` equation |
| `EMAX_i = EMAXREF + EMAXPS + EMAXCOMBO + EMAXHL + EMAXOTH + EMAXpedCNST + eta_EMAX` | n/a | Hu 2024 Results, nivolumab `EMAX_i` equation |
| `VC_i = VCREF * (LBM/55)^VCLBM * exp(VCSEX) * exp(VCado) * exp(VCped) * exp(eta_VC)` | n/a | Hu 2024 Results, nivolumab `VC_i` equation |
| `Q_i = QREF * (WTB/75)^QWTB * exp(eta_Q)`; `VP_i = VPREF * (LBM/55)^VPLBM * exp(eta_VP)` | n/a | Hu 2024 Results, nivolumab `Q_i` and `VP_i` equations |
| nivolumab `CL0REF` | 9.66 mL/h | Table 4 |
| nivolumab `VCREF`, `QREF`, `VPREF` | 4.01 L, 35.9 mL/h, 2.77 L | Table 4 |
| nivolumab `CLWTB`, `CLeGFR`, `VCLBM` | 0.630, 0.0982, 0.932 | Table 4 |
| nivolumab `CLSEX`, `CLPS`, `CLRAAA`, `CLRAAS` | -0.0998, 0.166, 0.0693, 0.00333 | Table 4 |
| nivolumab `EMAXREF`, `T50`, `HILL` | -0.298, 2670 h, 2.32 | Table 4 |
| nivolumab `CLHL`, `CLGBM`, `CLOTH` | -0.382, -0.578, 0.00699 | Table 4 |
| nivolumab `CLPEDST`, `CLADOST`, `CLPEDHL`, `CLPEDCNST` | -0.580, -0.223, -0.411, -0.801 | Table 4 |
| nivolumab `CLI1Q3`, `CLI3Q3`, `CLBVCO` | 0.0973, 0.349, 0.132 | Table 4 |
| nivolumab `EMAXPS`, `EMAXIPICO`, `EMAXHL`, `EMAXOTH`, `EMAXPEDCNST` | -0.157, -0.124, 0.132, 0.118, 0.696 | Table 4 |
| nivolumab `VCSEX`, `VCPED`, `VCADO` | 0.0195, -0.277, -0.273 | Table 4 |
| nivolumab omega^2 CL / VC / covariance / Emax | 0.108, 0.0751, 0.0220, 0.160 | Table 4 |
| nivolumab omega^2 Q / VP / covariance | constrained equal to the CL / VC block | Appendix File S2, `$OMEGA BLOCK(2) SAME` |
| nivolumab proportional residual error | 0.199 | Table 4 |
| `QWTB = CLWTB`, `VPLBM = VCLBM` | 0.630, 0.932 | Figure 1 caption ("fixed to be similar to those of CL and VC") |
| adult-CNS-tumour Emax = 0 (stationary CL) | n/a | Appendix File S2, `IF (POP_I .EQ. 4) EMAX = 0` |
| pediatric ST / HL Emax offsets = 0 | n/a | Methods; Appendix File S2 `THETA(38)` and `THETA(39)` = `0 FIX` |
| `CL_i = CL0TV_i * exp(eta_CL)` (ipilimumab, stationary) | n/a | Hu 2024 Results, first ipilimumab equation |
| `CL0TV_i = CL0REF * (LBM/55)^CLLBM * exp(...)` | n/a | Hu 2024 Results, ipilimumab `CL0TV,i` equation (see Errata on the printed subscript) |
| `VC_i = VCREF * (LBM/55)^VCLBM * exp(VCado) * exp(VCped) * exp(eta_VC)` | n/a | Hu 2024 Results, ipilimumab `VC_i` equation |
| ipilimumab `CLREF`, `VCREF`, `QREF`, `VPREF` | 13.5 mL/h, 3.90 L, 35.8 mL/h, 3.47 L | Table 5 |
| ipilimumab `CLLBM`, `VCLBM` | 0.789, 0.874 | Table 5 |
| ipilimumab `CLCNST`, `CLOTH`, `CLPEDOTH`, `CLPEDCNST`, `CLPEDMEL` | -0.661, -0.698, -0.462, -0.668, -0.347 | Table 5 |
| ipilimumab `CLN1Q3`, `CLN3Q3` | 0.0417, 0.316 | Table 5 |
| ipilimumab `VCPED`, `VCADO` | -0.296, -0.217 | Table 5 |
| ipilimumab omega^2 CL / VC / covariance | 0.147, 0.0531, 0.0258 | Table 5 |
| ipilimumab residual error (proportional, additive) | 0.185, 1.14 ug/mL | Table 5 |
| age bands `< 12`, `12-17`, `>= 18` years | n/a | Methods; Appendix Files S2 and S3 (`AGE <= 11`, `12 <= AGE <= 17`) |

# Structural verification of the covariate wiring

The paper reports every patient-population effect twice: once as a coefficient
in Table 4 / Table 5, and once as a percentage change in the Results narrative.
Because those percentages are exactly `exp(theta)`, reproducing them from a
typical-value solve is a direct check that each coefficient reaches the correct
(tumour type x age band) cell of the model -- the one thing a transcription of
this model can most easily get wrong.

```{r ref-cells}
# Reference covariates from Hu 2024 Table 4 Note / Results: 75 kg, LBM 55 kg,
# eGFR 90 mL/min/1.73 m^2, male, White/Other, performance status 0, monotherapy.
nivo_cells <- tibble::tribble(
  ~cell,                                   ~AGE, ~TUMTP_LYMPH, ~TUMTP_CNS_PRIM, ~TUMTP_OTHER,
  "Adult melanoma (reference)",              55,            0,               0,            0,
  "Adult lymphoma",                          36,            1,               0,            0,
  "Adult CNS tumour (glioblastoma)",         57,            0,               1,            0,
  "Adult other solid tumour",                62,            0,               0,            1,
  "Adolescent (12-17 y) solid tumour",       15,            0,               0,            1,
  "Young pediatric (<12 y) solid tumour",     8,            0,               0,            1,
  "Pediatric (<18 y) lymphoma",              15,            1,               0,            0,
  "Pediatric (<18 y) CNS tumour",            10,            0,               1,            0
) |>
  mutate(
    id = row_number(), WT = 75, LBM = 55, CRCL = 90, SEXF = 0,
    ECOG_GE1 = 0, RACE_BLACK = 0, RACE_ASIAN = 0,
    CONMED_IPI_1Q3W = 0, CONMED_IPI_3Q3W = 0,
    CONMED_BRENTUXIMAB = 0, CONMED_IPI_ANY = 0
  )

# One dose plus a t = 0 observation. At t = 0 the sigmoid time term is exactly
# zero, so the returned `cl` is the baseline CL0 the paper tabulates.
nivo_cell_ev <- bind_rows(
  nivo_cells |> mutate(time = 0, amt = 240, evid = 1L, cmt = "central"),
  nivo_cells |> mutate(time = 0, amt = NA_real_, evid = 0L, cmt = "central")
) |>
  arrange(id, desc(evid))

nivo_cell_sim <- rxode2::rxSolve(
  rxode2::zeroRe(nivo), events = nivo_cell_ev, keep = "cell"
) |>
  as.data.frame() |>
  group_by(cell) |>
  summarise(cl = first(cl), vc = first(vc), .groups = "drop")

nivo_ref <- nivo_cell_sim$cl[nivo_cell_sim$cell == "Adult melanoma (reference)"]
nivo_ref_vc <- nivo_cell_sim$vc[nivo_cell_sim$cell == "Adult melanoma (reference)"]
```

```{r ref-cells-table}
nivo_expected <- tibble::tribble(
  ~cell,                                   ~theta,   ~published_pct_lower,
  "Adult melanoma (reference)",             0,       0,
  "Adult lymphoma",                        -0.382,  32,
  "Adult CNS tumour (glioblastoma)",       -0.578,  44,
  "Adult other solid tumour",               0.00699, 0,
  "Adolescent (12-17 y) solid tumour",     -0.223,  20,
  "Young pediatric (<12 y) solid tumour",  -0.580,  44,
  "Pediatric (<18 y) lymphoma",            -0.411,  34,
  "Pediatric (<18 y) CNS tumour",          -0.801,  55
)

nivo_chk <- nivo_cell_sim |>
  left_join(nivo_expected, by = "cell") |>
  mutate(
    ratio_model    = cl / nivo_ref,
    ratio_expected = exp(theta),
    pct_lower      = 100 * (1 - ratio_model),
    abs_err        = abs(ratio_model - ratio_expected)
  )

# Deterministic (zeroRe) solve of a closed-form algebraic expression: the only
# error here is floating point, so the bound is tight on purpose.
stopifnot(max(nivo_chk$abs_err) < 1e-8)
# The narrative percentages in Hu 2024 Results are rounded to whole percent.
stopifnot(max(abs(nivo_chk$pct_lower - nivo_chk$published_pct_lower)) < 1)

nivo_chk |>
  select(cell, ratio_model, pct_lower, published_pct_lower) |>
  mutate(across(c(ratio_model, pct_lower), \(x) round(x, 3))) |>
  rename(
    "Patient population"                = cell,
    "CL0 / CL0 (adult MEL)"             = ratio_model,
    "% lower (model)"                   = pct_lower,
    "% lower (Hu 2024 Results)"         = published_pct_lower
  ) |>
  knitr::kable(
    caption = paste(
      "Nivolumab baseline-clearance covariate effects reproduced from the",
      "packaged model at the paper's reference covariates. Compare with the",
      "Hu 2024 Results narrative and Figure 1a."
    )
  )
```

The central-volume age effects are gated the same way. Hu 2024 reports both
pediatric bands at 24% lower `VC` than adults.

```{r vc-check}
nivo_vc_chk <- nivo_cell_sim |>
  mutate(
    band = case_when(
      cell %in% c("Adolescent (12-17 y) solid tumour",
                  "Pediatric (<18 y) lymphoma")        ~ "Adolescent (12-17 y)",
      cell %in% c("Young pediatric (<12 y) solid tumour",
                  "Pediatric (<18 y) CNS tumour")      ~ "Young pediatric (<12 y)",
      TRUE                                              ~ "Adult"
    ),
    ratio_vc = vc / nivo_ref_vc
  ) |>
  group_by(band) |>
  summarise(ratio_vc = mean(ratio_vc), .groups = "drop")

vc_expected <- c(
  "Adult"                   = 1,
  "Adolescent (12-17 y)"    = exp(-0.273),
  "Young pediatric (<12 y)" = exp(-0.277)
)
stopifnot(max(abs(nivo_vc_chk$ratio_vc - vc_expected[nivo_vc_chk$band])) < 1e-8)
# Hu 2024 Results: "adolescent (12-17 years) and pediatric (<12 years) patients
# had 24% lower VC than adult patients".
stopifnot(all(abs(100 * (1 - nivo_vc_chk$ratio_vc[nivo_vc_chk$band != "Adult"]) - 24) < 1))

nivo_vc_chk |>
  mutate(`% lower than adult` = round(100 * (1 - ratio_vc), 1),
         ratio_vc = round(ratio_vc, 3)) |>
  rename("Age band" = band, "VC / VC (adult)" = ratio_vc) |>
  knitr::kable(caption = "Nivolumab central-volume age effects (Hu 2024 Figure 1b).")
```

The same check for the ipilimumab model, whose patient-population factor pools
the whole under-18 range on `CL` (no 12-year split) but does split `VC` at 12
years.

```{r ipi-ref-cells}
ipi_cells <- tibble::tribble(
  ~cell,                                 ~AGE, ~TUMTP_MEL, ~TUMTP_CNS_PRIM, ~TUMTP_OTHER, ~theta, ~published_pct_lower,
  "Adult melanoma (reference)",            55,          1,               0,            0,  0,      0,
  "Adult CNS tumour",                      19,          0,               1,            0, -0.661, 48,
  "Adult other solid tumour",              21,          0,               0,            1, -0.698, 50,
  "Pediatric (<18 y) other solid tumour",  13,          0,               0,            1, -0.462, 37,
  "Pediatric (<18 y) CNS tumour",          10,          0,               1,            0, -0.668, 49,
  "Pediatric (<18 y) melanoma",            13,          1,               0,            0, -0.347, 29
) |>
  mutate(id = row_number(), LBM = 55, CONMED_NIVO_1Q3W = 0, CONMED_NIVO_3Q3W = 0)

ipi_cell_ev <- bind_rows(
  ipi_cells |> mutate(time = 0, amt = 200, evid = 1L, cmt = "central"),
  ipi_cells |> mutate(time = 0, amt = NA_real_, evid = 0L, cmt = "central")
) |>
  select(-theta, -published_pct_lower) |>
  arrange(id, desc(evid))

ipi_cell_sim <- rxode2::rxSolve(
  rxode2::zeroRe(ipi), events = ipi_cell_ev, keep = "cell"
) |>
  as.data.frame() |>
  group_by(cell) |>
  summarise(cl = first(cl), vc = first(vc), .groups = "drop")

ipi_ref <- ipi_cell_sim$cl[ipi_cell_sim$cell == "Adult melanoma (reference)"]

ipi_chk <- ipi_cell_sim |>
  left_join(ipi_cells |> select(cell, theta, published_pct_lower), by = "cell") |>
  mutate(ratio_model = cl / ipi_ref,
         pct_lower   = 100 * (1 - ratio_model),
         abs_err     = abs(ratio_model - exp(theta)))

stopifnot(max(ipi_chk$abs_err) < 1e-8)
stopifnot(max(abs(ipi_chk$pct_lower - ipi_chk$published_pct_lower)) < 1)

ipi_chk |>
  select(cell, ratio_model, pct_lower, published_pct_lower) |>
  mutate(across(c(ratio_model, pct_lower), \(x) round(x, 3))) |>
  rename(
    "Patient population"        = cell,
    "CL / CL (adult MEL)"       = ratio_model,
    "% lower (model)"           = pct_lower,
    "% lower (Hu 2024 Results)" = published_pct_lower
  ) |>
  knitr::kable(
    caption = paste(
      "Ipilimumab clearance covariate effects reproduced from the packaged",
      "model at LBM = 55 kg. Compare with the Hu 2024 Results narrative and",
      "Figure 2."
    )
  )
```

# The time-varying-clearance arm

Figure 1b of Hu 2024 reports the steady-state-to-baseline clearance ratio and
states the identity used to compute it: `CLss / CL0 = exp(EMAX)`. Simulating a
typical subject far past `T50` recovers that ratio, and does so per covariate
cell -- including the adult CNS-tumour cell, where the source control stream
sets `EMAX` to exactly zero so clearance is stationary. That last row is not
visible anywhere in the printed equations; it comes from Appendix File S2.

```{r emax-check}
emax_cells <- tibble::tribble(
  ~cell,                              ~AGE, ~TUMTP_LYMPH, ~TUMTP_CNS_PRIM, ~TUMTP_OTHER, ~ECOG_GE1, ~CONMED_IPI_ANY, ~emax_expected,
  "Adult MEL, mono, PS 0",              55,            0,               0,            0,         0,               0, -0.298,
  "Adult MEL, mono, PS > 0",            55,            0,               0,            0,         1,               0, -0.298 - 0.157,
  "Adult MEL + ipilimumab, PS 0",       55,            0,               0,            0,         0,               1, -0.298 - 0.124,
  "Adult lymphoma, mono, PS 0",         36,            1,               0,            0,         0,               0, -0.298 + 0.132,
  "Adult other ST, mono, PS 0",         62,            0,               0,            1,         0,               0, -0.298 + 0.118,
  "Pediatric CNS tumour, mono, PS 0",   10,            0,               1,            0,         0,               0, -0.298 + 0.696,
  "Adult CNS tumour (GBM), mono, PS 0", 57,            0,               1,            0,         0,               0,  0
) |>
  mutate(
    id = row_number(), WT = 75, LBM = 55, CRCL = 90, SEXF = 0,
    RACE_BLACK = 0, RACE_ASIAN = 0, CONMED_IPI_1Q3W = 0,
    CONMED_IPI_3Q3W = 0, CONMED_BRENTUXIMAB = 0
  )

# t = 0 gives CL0; t = 2000 days is ~18 x T50, so the sigmoid has reached
# 99.88% of its plateau. The residual approach error is < 0.1% and is the
# reason the tolerance below is 0.5% rather than machine epsilon.
emax_ev <- bind_rows(
  emax_cells |> mutate(time = 0, amt = 240, evid = 1L, cmt = "central"),
  tidyr::crossing(emax_cells, time = c(0, 2000)) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
) |>
  select(-emax_expected) |>
  arrange(id, time, desc(evid))

emax_sim <- rxode2::rxSolve(
  rxode2::zeroRe(nivo), events = emax_ev, keep = "cell"
) |>
  as.data.frame() |>
  group_by(cell) |>
  summarise(cl0 = cl[which.min(time)], clss = cl[which.max(time)], .groups = "drop") |>
  left_join(emax_cells |> select(cell, emax_expected), by = "cell") |>
  mutate(ratio_model = clss / cl0, ratio_expected = exp(emax_expected))

stopifnot(max(abs(emax_sim$ratio_model / emax_sim$ratio_expected - 1)) < 0.005)
# The stationary-CL cell must be exactly flat, not merely close.
stopifnot(abs(emax_sim$ratio_model[emax_sim$cell == "Adult CNS tumour (GBM), mono, PS 0"] - 1) < 1e-10)

emax_sim |>
  mutate(across(c(ratio_model, ratio_expected), \(x) round(x, 3))) |>
  select(cell, ratio_model, ratio_expected) |>
  rename(
    "Covariate cell"            = cell,
    "CLss / CL0 (model)"        = ratio_model,
    "exp(EMAX) (Hu 2024)"       = ratio_expected
  ) |>
  knitr::kable(
    caption = paste(
      "Nivolumab steady-state-to-baseline clearance ratio, reproducing the",
      "CLss/CL0 = exp(EMAX) identity of Hu 2024 Figure 1b. The last row is the",
      "stationary-clearance adult CNS-tumour cell defined only in Appendix",
      "File S2."
    )
  )
```

```{r cl-time-profile, fig.cap = "Time course of typical nivolumab clearance by patient population, reproducing the direction and magnitude of the CLss/CL0 effects in Hu 2024 Figure 1b."}
prof_ev <- bind_rows(
  emax_cells |> mutate(time = 0, amt = 240, evid = 1L, cmt = "central"),
  tidyr::crossing(emax_cells, time = seq(0, 700, by = 7)) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
) |>
  select(-emax_expected) |>
  arrange(id, time, desc(evid))

rxode2::rxSolve(rxode2::zeroRe(nivo), events = prof_ev, keep = "cell") |>
  as.data.frame() |>
  group_by(cell) |>
  mutate(rel_cl = cl / first(cl)) |>
  ungroup() |>
  ggplot(aes(time, rel_cl, colour = cell)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time since first dose (days)", y = expression(CL(t)/CL[0]),
       colour = NULL,
       title = "Time-varying nivolumab clearance by patient population") +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(ncol = 2))
```

# Virtual cohort

Individual patient data are not public. The cohorts below approximate the
covariate distributions of Hu 2024 Tables 1 and 2. Body weight is drawn from
the published per-population mean and SD, truncated to the published range;
lean body mass is then set to the published population mean `LBM/WT` ratio
times the drawn weight, with a small lognormal residual. Hu 2024 computed lean
body mass by the Boer and Peter equations, which require height -- a variable
the paper does not tabulate -- so this ratio construction is the closest
paper-sourced substitute (see Assumptions).

```{r cohort}
# set.seed() seeds R's RNG, not rxode2's; rxode2 partitions its streams per
# solver thread, so the cohort below differs between a 2-core CI runner and a
# 16-thread workstation. Every assertion downstream is written to hold for any
# cohort this model can produce.
set.seed(20240301)

n_arm <- 150L

draw_subjects <- function(n, wt_mean, wt_sd, wt_lo, wt_hi, lbm_ratio, id_offset) {
  wt <- pmin(pmax(rnorm(n, wt_mean, wt_sd), wt_lo), wt_hi)
  tibble(
    id  = id_offset + seq_len(n),
    WT  = wt,
    LBM = wt * lbm_ratio * exp(rnorm(n, 0, 0.06))
  )
}

# Hu 2024 Table 1: adult MEL weight 82.3 (18.1) kg, LBM 57.7 (10.8) kg;
# pediatric ST weight 44.1 (23.6) kg, LBM 35.4 (16.1) kg (median age 12 y).
adult_mel  <- \(n, off) draw_subjects(n, 82.3, 18.1, 40, 160, 57.7 / 82.3, off)
adol_st    <- \(n, off) draw_subjects(n, 55.0, 15.0, 30, 100, 35.4 / 44.1, off)
```

The adolescent-solid-tumour weight distribution is centred at 55 kg rather than
the whole pediatric-cohort mean of 44.1 kg, because the pediatric solid-tumour
row of Table 1 pools 1-17 year olds while the dosing question in Hu 2024
concerns 12-17 year olds specifically (Figure S2 shows the adolescent
weight-age relationship the authors sampled from NHANES).

```{r events}
# 30-minute IV infusions (see Assumptions).
INF_DUR <- 0.5 / 24

make_arm <- function(subjects, arm, dose_fun, dose_times, obs_times,
                     covariates) {
  subj <- subjects |> mutate(arm = arm) |> bind_cols(covariates)
  doses <- tidyr::crossing(subj, time = dose_times) |>
    mutate(amt = dose_fun(WT), evid = 1L, cmt = "central",
           rate = amt / INF_DUR)
  obs <- tidyr::crossing(subj, time = obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central", rate = 0)
  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}

nivo_cov_adult <- tibble(
  AGE = 55, CRCL = 90, SEXF = 0, ECOG_GE1 = 0, RACE_BLACK = 0, RACE_ASIAN = 0,
  TUMTP_LYMPH = 0, TUMTP_CNS_PRIM = 0, TUMTP_OTHER = 0,
  CONMED_IPI_1Q3W = 0, CONMED_IPI_3Q3W = 0, CONMED_BRENTUXIMAB = 0,
  CONMED_IPI_ANY = 0
)
nivo_cov_adol <- nivo_cov_adult |> mutate(AGE = 15, TUMTP_OTHER = 1)

# Twelve q2w doses reaches steady state for a mAb with a ~3-week half-life;
# NCA is taken over the last (twelfth) dosing interval.
nivo_dose_times <- seq(0, by = 14, length.out = 12)
nivo_obs_times  <- sort(unique(c(seq(0, 154, by = 14), seq(154, 168, by = 0.5))))

nivo_events <- bind_rows(
  make_arm(adult_mel(n_arm, 0L),      "Adult MEL, 240 mg q2w",
           \(wt) 240,                  nivo_dose_times, nivo_obs_times, nivo_cov_adult),
  make_arm(adol_st(n_arm, 1000L),     "Adolescent ST, 240 mg q2w",
           \(wt) 240,                  nivo_dose_times, nivo_obs_times, nivo_cov_adol),
  make_arm(adol_st(n_arm, 2000L),     "Adolescent ST, 3 mg/kg q2w (cap 240 mg)",
           \(wt) pmin(3 * wt, 240),    nivo_dose_times, nivo_obs_times, nivo_cov_adol)
)
stopifnot(!anyDuplicated(unique(nivo_events[, c("id", "time", "evid")])))

ipi_cov_adult <- tibble(
  AGE = 55, TUMTP_MEL = 1, TUMTP_CNS_PRIM = 0, TUMTP_OTHER = 0,
  CONMED_NIVO_1Q3W = 0, CONMED_NIVO_3Q3W = 0
)
ipi_cov_adol <- ipi_cov_adult |> mutate(AGE = 15)

# Approved adult ipilimumab melanoma regimen: 3 mg/kg q3w for four doses.
ipi_dose_times <- c(0, 21, 42, 63)
ipi_obs_times  <- sort(unique(c(seq(0, 63, by = 7), seq(63, 84, by = 0.5))))

ipi_events <- bind_rows(
  make_arm(adult_mel(n_arm, 0L),  "Adult MEL, 3 mg/kg q3w",
           \(wt) 3 * wt, ipi_dose_times, ipi_obs_times, ipi_cov_adult),
  make_arm(adol_st(n_arm, 1000L), "Adolescent MEL, 3 mg/kg q3w",
           \(wt) 3 * wt, ipi_dose_times, ipi_obs_times, ipi_cov_adol)
)
stopifnot(!anyDuplicated(unique(ipi_events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
nivo_sim <- rxode2::rxSolve(
  nivo, events = as.data.frame(nivo_events), keep = "arm"
) |>
  as.data.frame()

ipi_sim <- rxode2::rxSolve(
  ipi, events = as.data.frame(ipi_events), keep = "arm"
) |>
  as.data.frame()

stopifnot(nrow(nivo_sim) > 0, nrow(ipi_sim) > 0)
```

```{r figure-s9-s10, fig.cap = "Steady-state nivolumab concentration-time profiles over the twelfth dosing interval. Adolescents on the adult flat dose sit above adults; weight-based dosing with a 240 mg cap brings them back into the adult range. Reproduces the qualitative finding of Hu 2024 Figures S9 and S10."}
nivo_sim |>
  filter(time >= 154) |>
  group_by(arm, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time - 154, Q50, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time since twelfth dose (days)", y = "Nivolumab (ug/mL)",
       colour = NULL, fill = NULL,
       title = "Nivolumab at steady state: adolescents vs adults") +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(ncol = 1), fill = guide_legend(ncol = 1))
```

# PKNCA validation

Hu 2024 publishes no NCA table, so the NCA below serves two purposes: an
internal numerical check that the model's units are self-consistent, and a
quantitative statement of the exposure comparisons the paper makes only in
figures.

## Nivolumab: steady-state exposure over the last dosing interval

```{r pknca-nivo}
nivo_nca_conc <- nivo_sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, arm)

nivo_nca_conc <- bind_rows(
  nivo_nca_conc,
  nivo_nca_conc |> distinct(id, arm) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, arm, time, .keep_all = TRUE) |>
  arrange(id, arm, time)

nivo_conc_obj <- PKNCA::PKNCAconc(nivo_nca_conc, Cc ~ time | arm + id)

nivo_dose_obj <- PKNCA::PKNCAdose(
  nivo_events |> filter(evid == 1L) |> select(id, time, amt, arm),
  amt ~ time | arm + id
)

nivo_intervals <- data.frame(
  start = 154, end = 168,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, cav = TRUE, ctrough = TRUE
)

nivo_nca <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(nivo_conc_obj, nivo_dose_obj, intervals = nivo_intervals)
)

nivo_nca_tab <- as.data.frame(nivo_nca) |>
  filter(PPTESTCD %in% c("cmax", "cav", "auclast", "ctrough")) |>
  group_by(arm, PPTESTCD) |>
  summarise(gm = exp(mean(log(PPORRES))), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = gm)

nivo_nca_tab |>
  mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
  rename(
    "Arm"                          = arm,
    "Cavg,ss (ug/mL)"              = cav,
    "Cmax,ss (ug/mL)"              = cmax,
    "Ctrough,ss (ug/mL)"           = ctrough,
    "AUCtau,ss (ug*day/mL)"        = auclast
  ) |>
  knitr::kable(
    caption = paste(
      "Geometric-mean steady-state nivolumab exposure over the twelfth q2w",
      "dosing interval (n = 150 per arm)."
    )
  )
```

```{r pknca-nivo-claims}
gm <- \(a, p) nivo_nca_tab[[p]][nivo_nca_tab$arm == a]

flat_adult <- gm("Adult MEL, 240 mg q2w", "cav")
flat_adol  <- gm("Adolescent ST, 240 mg q2w", "cav")
wt_adol    <- gm("Adolescent ST, 3 mg/kg q2w (cap 240 mg)", "cav")

flat_ratio <- 100 * flat_adol / flat_adult
wt_ratio   <- 100 * wt_adol   / flat_adult

# Hu 2024: "Steady-state nivolumab exposures in adolescent patients with STs
# ... who received the approved adult flat dose of nivolumab monotherapy
# (240 mg q2w) were predicted to exceed those of adults" (Figure S9), whereas
# "With body-weight-based dosing at 3 mg/kg q2w (up to a maximum dose of
# 240 mg), nivolumab steady-state exposures in adolescents ... were predicted
# to be similar to those of adults" (Figure S10).
#
# These are large, structurally driven effects: the flat-dose arm is elevated
# by BOTH the ~20% lower adolescent solid-tumour CL and the smaller body size,
# while weight-based dosing removes the size component. The bounds below are
# deliberately wide of any single cohort draw -- an exceedance of "at least
# 20%" and a similarity window of "within 35%" -- while still going red on a
# mis-transcribed clearance coefficient or dose, which move these ratios by
# tens of percent.
stopifnot(flat_ratio > 120)
stopifnot(abs(wt_ratio - 100) < 35)
# Weight-based dosing must be the closer of the two to the adult reference.
stopifnot(abs(wt_ratio - 100) < abs(flat_ratio - 100) - 10)

tibble::tibble(
  Claim = c(
    "Adolescent ST on adult flat 240 mg q2w exceeds adult MEL exposure",
    "Adolescent ST on 3 mg/kg q2w (cap 240 mg) is similar to adult MEL"
  ),
  `Hu 2024 source` = c("Figure S9", "Figure S10"),
  `Cavg,ss vs adult (%)` = round(c(flat_ratio, wt_ratio), 1)
) |>
  knitr::kable(caption = "Nivolumab dosing claims reproduced from the packaged model.")
```

## Ipilimumab: fourth-dose exposure, and a closed-form clearance identity

Ipilimumab clearance is stationary in this model, so a single-dose `AUCinf`
must equal `Dose / CL` exactly. Because the typical-value solve and the
closed form use the same parameters, the only difference between them is
numerical integration error, and the bound below is correspondingly tight.
This is the check that catches a units slip in the `mL/h` to `L/day`
conversion.

```{r ipi-identity}
ident_subj <- tibble(
  id = 1L, LBM = 55, AGE = 55, TUMTP_MEL = 1, TUMTP_CNS_PRIM = 0,
  TUMTP_OTHER = 0, CONMED_NIVO_1Q3W = 0, CONMED_NIVO_3Q3W = 0
)
ident_dose <- 200

ident_ev <- bind_rows(
  ident_subj |> mutate(time = 0, amt = ident_dose, evid = 1L, cmt = "central",
                       rate = ident_dose / INF_DUR),
  tidyr::crossing(ident_subj, time = c(seq(0, 5, by = 0.05), seq(5, 400, by = 0.25))) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central", rate = NA_real_)
) |>
  arrange(time, desc(evid)) |>
  distinct(time, evid, .keep_all = TRUE)

ident_sim <- rxode2::rxSolve(rxode2::zeroRe(ipi), events = as.data.frame(ident_ev)) |>
  as.data.frame()
ident_cl <- ident_sim$cl[1]

ident_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(
    ident_sim |> filter(!is.na(Cc)) |> select(time, Cc) |> mutate(id = 1L),
    Cc ~ time | id
  ),
  PKNCA::PKNCAdose(
    data.frame(id = 1L, time = 0, amt = ident_dose), amt ~ time | id
  ),
  intervals = data.frame(start = 0, end = Inf, aucinf.obs = TRUE, half.life = TRUE)
))

ident_res  <- as.data.frame(ident_nca)
auc_nca    <- ident_res$PPORRES[ident_res$PPTESTCD == "aucinf.obs"]
auc_closed <- ident_dose / ident_cl   # mg / (L/day) = mg*day/L = ug*day/mL

stopifnot(abs(auc_nca / auc_closed - 1) < 0.01)

tibble::tibble(
  Quantity = c("CL (L/day)", "AUCinf from PKNCA (ug*day/mL)",
               "Dose / CL closed form (ug*day/mL)", "Relative difference (%)"),
  Value = signif(c(ident_cl, auc_nca, auc_closed,
                   100 * (auc_nca / auc_closed - 1)), 4)
) |>
  knitr::kable(
    caption = paste(
      "Ipilimumab single-dose mass-balance identity for a reference adult",
      "(LBM 55 kg, melanoma, monotherapy; 200 mg IV). Stationary clearance",
      "makes AUCinf = Dose/CL exact up to integration error."
    )
  )
```

Hu 2024 states that at the approved 3 mg/kg q3w regimen, adolescent ipilimumab
exposure after the fourth dose exceeds adult exposure (Figure S13). The check
below makes that comparison on *typical* subjects with `zeroRe()`, so it is a
deterministic function of the parameters: the only thing that can move it is a
mis-transcribed coefficient, not which subjects a cohort happened to draw. This
is the load-bearing gate on the claim; the stochastic cohort that follows is
for display.

```{r ipi-typical-exposure}
# Typical adult MEL (Table 2: WT 82.3 kg, LBM 57.7 kg) versus a typical
# adolescent MEL subject on the same mg/kg dose.
ipi_typ <- tibble::tribble(
  ~cell,                          ~AGE, ~WT,  ~LBM,
  "Adult MEL, 3 mg/kg q3w",         55, 82.3, 57.7,
  "Adolescent MEL, 3 mg/kg q3w",    15, 55.0, 55.0 * 35.4 / 44.1
) |>
  mutate(id = row_number(), TUMTP_MEL = 1, TUMTP_CNS_PRIM = 0, TUMTP_OTHER = 0,
         CONMED_NIVO_1Q3W = 0, CONMED_NIVO_3Q3W = 0)

ipi_typ_ev <- bind_rows(
  tidyr::crossing(ipi_typ, time = ipi_dose_times) |>
    mutate(amt = 3 * WT, evid = 1L, cmt = "central", rate = amt / INF_DUR),
  tidyr::crossing(ipi_typ, time = ipi_obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central", rate = 0)
) |>
  arrange(id, time, desc(evid))

ipi_typ_sim <- rxode2::rxSolve(
  rxode2::zeroRe(ipi), events = as.data.frame(ipi_typ_ev), keep = "cell"
) |>
  as.data.frame()

ipi_typ_conc <- ipi_typ_sim |> filter(!is.na(Cc)) |> select(id, time, Cc, cell)
ipi_typ_conc <- bind_rows(
  ipi_typ_conc,
  ipi_typ_conc |> distinct(id, cell) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, cell, time, .keep_all = TRUE) |>
  arrange(id, time)

ipi_typ_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(ipi_typ_conc, Cc ~ time | cell + id),
  PKNCA::PKNCAdose(
    ipi_typ_ev |> filter(evid == 1L) |> select(id, time, amt, cell),
    amt ~ time | cell + id
  ),
  intervals = data.frame(start = 63, end = 84, cav = TRUE)
))

ipi_typ_tab <- as.data.frame(ipi_typ_nca) |> filter(PPTESTCD == "cav")
ipi_typ_ratio <- 100 *
  ipi_typ_tab$PPORRES[ipi_typ_tab$cell == "Adolescent MEL, 3 mg/kg q3w"] /
  ipi_typ_tab$PPORRES[ipi_typ_tab$cell == "Adult MEL, 3 mg/kg q3w"]

# Deterministic, so this is bounded tightly. The value is driven by the 29%
# lower pediatric-melanoma CL (exp(-0.347)) partly offset by the smaller
# mg/kg dose that a lighter subject receives; a sign error on CL_PEDMEL or a
# mis-wired LBM exponent moves it by tens of percent.
stopifnot(ipi_typ_ratio > 100)
stopifnot(abs(ipi_typ_ratio - 113.6) < 3)

tibble::tibble(
  Quantity = c("Typical adult Cavg (ug/mL)", "Typical adolescent Cavg (ug/mL)",
               "Adolescent / adult (%)"),
  Value = signif(c(
    ipi_typ_tab$PPORRES[ipi_typ_tab$cell == "Adult MEL, 3 mg/kg q3w"],
    ipi_typ_tab$PPORRES[ipi_typ_tab$cell == "Adolescent MEL, 3 mg/kg q3w"],
    ipi_typ_ratio
  ), 4)
) |>
  knitr::kable(
    caption = paste(
      "Typical-value ipilimumab exposure over the fourth q3w dosing interval.",
      "Adolescent exposure exceeds adult, reproducing Hu 2024 Figure S13."
    )
  )
```

```{r pknca-ipi}
ipi_nca_conc <- ipi_sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, arm)

ipi_nca_conc <- bind_rows(
  ipi_nca_conc,
  ipi_nca_conc |> distinct(id, arm) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, arm, time, .keep_all = TRUE) |>
  arrange(id, arm, time)

ipi_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(ipi_nca_conc, Cc ~ time | arm + id),
  PKNCA::PKNCAdose(
    ipi_events |> filter(evid == 1L) |> select(id, time, amt, arm),
    amt ~ time | arm + id
  ),
  intervals = data.frame(
    start = 63, end = 84,
    cmax = TRUE, cav = TRUE, auclast = TRUE, ctrough = TRUE
  )
))

ipi_nca_tab <- as.data.frame(ipi_nca) |>
  filter(PPTESTCD %in% c("cmax", "cav", "auclast", "ctrough")) |>
  group_by(arm, PPTESTCD) |>
  summarise(gm = exp(mean(log(PPORRES))), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = gm)

ipi_ratio <- 100 *
  ipi_nca_tab$cav[ipi_nca_tab$arm == "Adolescent MEL, 3 mg/kg q3w"] /
  ipi_nca_tab$cav[ipi_nca_tab$arm == "Adult MEL, 3 mg/kg q3w"]

# This is the SAME comparison as the typical-value check above, but taken over
# a random cohort, and it is deliberately gated far more loosely. The
# geometric-mean ratio of a 150-per-arm draw moves over roughly 106-123% purely
# with the eta draw (omega^2_CL = 0.147 is large, and rxSetSeed() fixes
# rxode2's stream per solver thread, so CI draws a different cohort than a
# workstation does). Gating this near its own central value is exactly the
# assertion shape that passes locally and then fails CI. The transcription
# check lives in `ipi-typical-exposure`; this one only confirms the cohort
# lands on the same side of parity.
stopifnot(ipi_ratio > 100)

ipi_nca_tab |>
  mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
  rename(
    "Arm"                        = arm,
    "Cavg (ug/mL)"               = cav,
    "Cmax (ug/mL)"               = cmax,
    "Ctrough (ug/mL)"            = ctrough,
    "AUCtau (ug*day/mL)"         = auclast
  ) |>
  knitr::kable(
    caption = paste(
      "Geometric-mean ipilimumab exposure over the fourth q3w dosing interval",
      "(n = 150 per arm). Adolescent exposure is",
      paste0(round(ipi_ratio), "%"), "of adult, reproducing Hu 2024 Figure S13."
    )
  )
```

# Assumptions and deviations

- **Time units.** Hu 2024 reports `CL` and `Q` in mL/h and `T50` in hours. Both
  model files convert to L/day and days (`x 24 / 1000` and `/ 24`) so the time
  axis matches the sibling nivolumab and ipilimumab models in nlmixr2lib
  (`Bajaj_2017_nivolumab`, `Zhang_2019_nivolumab`, `Feng_2019_ipilimumab`).
  No value is otherwise altered.

- **`cl_time_max` is not log-transformed.** `checkModelConventions()` suggests
  renaming the `Emax` parameter to `lcl_time_max` and back-transforming. That
  is not possible here: `EMAXREF = -0.298` is negative by construction (it is
  the log-scale maximal *decrease* in clearance), and Hu 2024 places a normally
  distributed additive eta on it. The model therefore keeps the linear-scale
  name `cl_time_max`, matching `Bajaj_2017_nivolumab` and
  `Zhang_2019_nivolumab`.

- **Patient population is encoded as tumour-type indicators crossed with age
  bands.** Hu 2024 uses a single multi-level `POPN` column. Its levels are a
  cross of tumour type and age band (adult MEL, adult HL, adult GBM, adult
  other ST, pediatric ST, pediatric HL, pediatric CNST), so the model files
  reconstruct them from the canonical `TUMTP_MEL` / `TUMTP_LYMPH` /
  `TUMTP_CNS_PRIM` / `TUMTP_OTHER` indicators and `AGE`. One consequence is
  worth noting: the analysis dataset assigns its single pediatric melanoma
  subject to the pediatric solid-tumour group (Table 1 footnote d), so a
  pediatric `TUMTP_MEL = 1` record does not occur in the nivolumab source data
  even though the encoding admits one. The companion ipilimumab model does
  estimate a pediatric-melanoma effect.

- **Structural details taken from the supplement, not the printed equations.**
  Three facts that materially change the model are in the NONMEM control
  streams (Appendix Files S2 and S3) and nowhere in the article body:
  (i) adult CNS-tumour subjects have `EMAX` set to exactly zero, so their
  clearance is stationary; (ii) the nivolumab `Q` / `VP` random effects are
  declared `$OMEGA BLOCK(2) SAME`, i.e. constrained to the same variances and
  covariance as the `CL` / `VC` block -- which is why Table 4 lists no
  `omega^2` for them; (iii) the ipilimumab model has no `Q` or `VP` random
  effects at all. Without the supplement, (i) would be missed entirely and (ii)
  would have had to be guessed at.

- **A printed-equation subscript in the ipilimumab model is a typo.** The
  ipilimumab `CL0TV,i` equation prints the lean-body-mass power as `CLWTB`,
  but the base of the power is `LBM_i / LBM_REF`, Table 5 has no `CLWTB` row
  (only `CL_LBM = 0.789`), the surrounding prose says "the typical CL value at
  the reference value of LBM (55 kg)", and Appendix File S3 codes
  `CL_LBM = (LBM_I/LBM_R)**THETA(7)`. The model uses `CL_LBM = 0.789`. The
  label appears to be carried over from the body-weight-based base model.

- **Zero-fixed covariate coefficients are recorded, not modelled.** The
  nivolumab dataset contains ipilimumab 1 mg/kg q6w, radiotherapy, and
  radiotherapy-plus-temozolomide arms, and the ipilimumab dataset an oral
  budesonide arm, whose clearance coefficients were declared `0 FIX` in the
  control streams and do not appear in Tables 4 or 5. They are documented in
  each model file's `covariatesDataExcluded` list rather than given a
  `fixed(0)` parameter, since a coefficient held at zero contributes no term to
  the equations. The numeric power effects of age on `CL` (both models) and on
  `VC` (ipilimumab) were likewise `0 FIX`; age enters both models only through
  the categorical bands.

- **Lean body mass in the virtual cohort.** Hu 2024 computed lean body mass by
  the Boer equation (adults) and the Peter equation (children), both of which
  need height. The paper tabulates weight and lean body mass but not height, so
  the cohort sets `LBM = (published LBM mean / published WT mean) x WT` with a
  6% lognormal residual. This reproduces the published per-population mean
  `LBM/WT` ratio exactly and is used only for the exposure-comparison figures;
  the covariate-wiring and clearance-identity checks above all run at the
  paper's own reference covariates and do not depend on it.

- **Infusion duration.** The analysis datasets carry a `DOSEDUR` column but the
  paper does not report infusion durations. The simulations use 30-minute IV
  infusions. This affects `Cmax` but not `AUC`, `Cavg`, or `Ctrough`, and none
  of the gated comparisons depend on it.

- **Adolescent weight distribution.** Hu 2024 simulated adolescents by sampling
  the 2017-2018 NHANES database (Figure S2), which is not redistributed here.
  The adolescent arm draws weight from a normal distribution centred at 55 kg
  (SD 15, truncated to 30-100 kg), chosen to sit inside the published
  adolescent range rather than the 1-17 year pooled pediatric mean of 44.1 kg.

- **The third model in the paper is not extracted.** Appendix S1 reports a
  separate nivolumab population PK model (Full5) for adolescents receiving
  adjuvant melanoma treatment, fitted to a different dataset (24,546
  concentrations, 3,965 patients, 11 studies). Its covariate effects are given
  as percentages in Figure S17, but none of its structural parameters --
  `CL0REF`, `VCREF`, `QREF`, `VPREF`, `EMAXREF`, `T50`, `HILL`, the random
  effects, or the residual error -- is reported anywhere in the article, the
  supplement text, or the figure panels. Figure S17 panels A and B were both
  inspected directly at publisher resolution and carry only covariate ratios.
  The model is therefore not reconstructable and is omitted; the two models in
  the article body (Full1c for nivolumab, Full1 for ipilimumab) are fully
  specified and are what this vignette validates.

- **No published NCA to compare against.** Hu 2024 reports no Cmax / AUC /
  half-life table, so there is no `ncaComparisonTable()` block here. The NCA
  section instead gates on the closed-form `AUCinf = Dose/CL` identity for the
  stationary ipilimumab model and on the three exposure comparisons the paper
  states in prose for Figures S9, S10, and S13.
