---
title: "Tacrolimus (Huang 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tacrolimus (Huang 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Huang_2022_tacrolimus"))$meta$reference`
* Description: `r rxode2::rxode(readModelDb("Huang_2022_tacrolimus"))$meta$description`
* Article: <https://doi.org/10.3389/fphar.2022.942129> (open access, PMC9706003)

Huang et al. (2022) fitted a population PK model to routine whole-blood
tacrolimus concentrations from children with refractory nephrotic syndrome in
Phoenix NLME (FOCE-ELS), then fed the retained covariates to six machine-learning
regressions of the individual clearance. Only the population PK model is
packaged here; the machine-learning models (Table 4) are regressions on the
Bayesian clearance estimates and are not reported in a form that can be
reproduced. The final model (Results, Eqs. 1-2) is

* V/F (L) = 192.03 x exp(eta_V)
* CL/F (L/h) = 10.54 x (AGE/5.3)^0.31 x exp(-0.34 x CTLA4 GA) x
  exp(-0.15 x CTLA4 AA) x exp(-0.25 x CYP3A5\*3/\*3) x exp(-0.34 x combWZ) x
  exp(eta_CL)
* Ka = 4.48 1/h (fixed)

where CTLA4 GA and CTLA4 AA are indicators of the rs4553808 genotype
(reference GG), CYP3A5\*3/\*3 = 1 for a CYP3A5 non-expresser, and combWZ = 1 for
concomitant Wuzhi capsule (a *Schisandra sphenanthera* extract that inhibits
CYP3A and P-glycoprotein).

# Population

139 Chinese children treated at Guangzhou Women and Children's Medical Center
between August 2013 and December 2018 (Table 1): age median 5.3 (range
1.1-15.6) years, weight median 19.7 (9.5-88) kg, serum albumin median 18.3
(7.6-46.8) g/L. Table 1 prints gender as 37 male / 102 female. All had
steroid-resistant or steroid-dependent nephrotic syndrome and took oral
tacrolimus (Prograf) with prednisone or prednisolone; 6 also took Wuzhi
capsule. Starting doses were 500-3000 ug every 12 h, titrated to a trough of
5-10 ug/L. The 432 concentrations are mostly pre-dose troughs, with 35 peak
samples at 0.5-3 h post-dose; median concentration 6.48 (2.0-41.2) ug/L.
Genotypes (Table 2): CYP3A5 \*1/\*1 18, \*1/\*3 58, \*3/\*3 63; CTLA4 rs4553808
GG 2, GA 29, AA 108.

# Source trace

| Element | Value | Source |
|---|---|---|
| Structure: 1-compartment, first-order absorption and elimination | -- | Methods, Base model; Results |
| `lka` | 4.48 1/h (fixed) | Methods, Base model |
| `lcl` | 10.54 L/h | Table 3 (tvCL); Eq. 2 |
| `lvc` | 192.03 L | Table 3 (tvV); Eq. 1 |
| `e_age_cl` | 0.31, form `(AGE/5.3)^theta` | Table 3 (dCLdAGE); Eq. 2; Methods covariate form |
| `e_ctla4_ga_cl` | -0.34, form `exp(theta * GA)` | Table 3 (dCLdCTLA4 GA); Eq. 2 |
| `e_ctla4_aa_cl` | -0.15, form `exp(theta * AA)` | Table 3 (dCldCTLA4 AA); Eq. 2 |
| `e_cyp3a5_expr_cl` | -0.25, form `exp(theta * (1 - CYP3A5_EXPR))` | Table 3 (dCLdCYP3A5 \*3/\*3); Eq. 2 |
| `e_conmed_wuzhi_cl` | -0.34, form `exp(theta * combWZ)` | Table 3 (dCLdCombWZ); Eq. 2 |
| Exponential IIV `P = TV * exp(eta)` | -- | Methods, Base model |
| `etalvc` | 1.13 (variance) | Table 3 (omega2 V, shrinkage 38.09%) |
| `etalcl` | 0.13 (variance) | Table 3 (omega2 CL, shrinkage 30.02%) |
| Residual error `Yobs = Ypred * (1 + eps)` | -- | Methods; Results (proportional chosen) |
| `propSd` | 0.2995 | Table 3 (sigma = 29.95%; bootstrap median 0.30) |
| Covariate coding | GA, AA, \*3/\*3, combWZ = 1 when present, else 0 | Text below Eq. 2 |

# Typical CL/F by covariate group

Eq. 2 fixes the typical CL/F of every covariate combination. The model's
typical-value `cl` must reproduce it exactly.

```{r typical-cl}
mod <- readModelDb("Huang_2022_tacrolimus")
mod_typ <- rxode2::zeroRe(mod)

grid <- tidyr::expand_grid(
  ctla4 = c("GG", "GA", "AA"),
  CYP3A5_EXPR = c(1, 0),
  CONMED_WUZHI = c(0, 1),
  AGE = c(2, 5.3, 12)
) |>
  mutate(
    SNP_CTLA4_RS4553808_GA = as.numeric(ctla4 == "GA"),
    SNP_CTLA4_RS4553808_AA = as.numeric(ctla4 == "AA"),
    id = row_number(), time = 1, amt = 0, evid = 0L, cmt = "central"
  )
cl_typ <- rxode2::rxSolve(
  mod_typ,
  events = grid,
  keep = c("ctla4", "CYP3A5_EXPR", "CONMED_WUZHI", "AGE")
) |>
  as.data.frame() |>
  mutate(
    cl_eq2 = 10.54 * (AGE / 5.3)^0.31 *
      exp(-0.34 * (ctla4 == "GA")) * exp(-0.15 * (ctla4 == "AA")) *
      exp(-0.25 * (1 - CYP3A5_EXPR)) * exp(-0.34 * CONMED_WUZHI)
  )

cl_typ |>
  filter(AGE == 5.3, CONMED_WUZHI == 0) |>
  mutate(cyp3a5 = ifelse(CYP3A5_EXPR == 1, "expresser", "non-expresser")) |>
  select(ctla4, cyp3a5, cl, cl_eq2) |>
  dplyr::rename(
    "CTLA4 rs4553808" = ctla4, "CYP3A5" = cyp3a5,
    "Model CL/F (L/h)" = cl, "Eq. 2 CL/F (L/h)" = cl_eq2
  ) |>
  knitr::kable(digits = 3, caption = "Typical CL/F at 5.3 years without Wuzhi capsule.")

# Same typical parameters on both sides; the difference is floating-point only.
stopifnot(nrow(cl_typ) == 36, max(abs(cl_typ$cl / cl_typ$cl_eq2 - 1)) < 1e-10)
```

The CTLA4 effect is not monotone in allele count: GA patients clear
tacrolimus 29% slower than the GG reference (`exp(-0.34)` = 0.71) and AA
patients 14% slower (`exp(-0.15)` = 0.86). The GG reference group holds only
two patients, so the typical value 10.54 L/h describes a genotype that is
rare in this cohort; the typical CL/F of the most common combination (AA,
CYP3A5\*3/\*3, 5.3 years, no Wuzhi capsule) is
`r round(10.54 * exp(-0.15) * exp(-0.25), 2)` L/h, close to the 7.05 L/h of
the covariate-free base model reported in the Results.

```{r age-curve, fig.cap = "Typical CL/F against age for each CTLA4 genotype and CYP3A5 status, without Wuzhi capsule (Eq. 2)."}
age_curve <- tidyr::expand_grid(
  ctla4 = c("GG", "GA", "AA"),
  CYP3A5_EXPR = c(1, 0),
  AGE = seq(1, 16, by = 0.25)
) |>
  mutate(
    SNP_CTLA4_RS4553808_GA = as.numeric(ctla4 == "GA"),
    SNP_CTLA4_RS4553808_AA = as.numeric(ctla4 == "AA"),
    CONMED_WUZHI = 0,
    cyp3a5 = ifelse(CYP3A5_EXPR == 1, "CYP3A5 expresser", "CYP3A5 non-expresser"),
    id = row_number(), time = 1, amt = 0, evid = 0L, cmt = "central"
  )
rxode2::rxSolve(mod_typ, events = age_curve, keep = c("ctla4", "cyp3a5", "AGE")) |>
  as.data.frame() |>
  ggplot(aes(AGE, cl, colour = ctla4)) +
  geom_line() +
  facet_wrap(~cyp3a5) +
  labs(x = "Age (years)", y = "Typical CL/F (L/h)", colour = "CTLA4 rs4553808")
```

# Dose consistency with the observed troughs

The paper does not tabulate individual doses, so the observed concentrations
cannot be compared with a simulation directly. As a structural check, the
steady-state trough of the typical patient (5.3 years, CTLA4 AA, CYP3A5\*3/\*3,
no Wuzhi capsule) on 1 mg every 12 h gives the twice-daily dose that patient
would need to sit at the cohort's median concentration of 6.48 ug/L. A
mis-transcribed CL/F, V/F or unit would move that dose far outside the
500-3000 ug starting range the Methods report.

```{r dose-check}
tau <- 12
ndose <- 42 # 21 days twice daily, > 25 typical half-lives
t_last <- (ndose - 1) * tau

typ_pt <- data.frame(
  AGE = 5.3, SNP_CTLA4_RS4553808_GA = 0, SNP_CTLA4_RS4553808_AA = 1,
  CYP3A5_EXPR = 0, CONMED_WUZHI = 0
)
ev_typ <- bind_rows(
  data.frame(id = 1L, time = seq(0, by = tau, length.out = ndose), amt = 1, evid = 1L, cmt = "depot"),
  data.frame(id = 1L, time = t_last + tau, amt = 0, evid = 0L, cmt = "central")
) |>
  cbind(typ_pt)
trough_1mg <- rxode2::rxSolve(mod_typ, events = ev_typ) |>
  as.data.frame() |>
  pull(Cc)
dose_needed <- 6.48 / trough_1mg

tibble(
  quantity = c("Steady-state trough on 1 mg q12h (ng/mL)", "Dose for a 6.48 ng/mL trough (mg q12h)"),
  value = c(trough_1mg, dose_needed)
) |>
  dplyr::rename("Quantity" = quantity, "Value" = value) |>
  knitr::kable(digits = 2)

stopifnot(dose_needed >= 0.5, dose_needed <= 3)
```

# Virtual cohort and concentration-time profile

A cohort of 200 children is simulated with ages drawn log-normally around the
5.3-year median and truncated to the observed 1.1-15.6 years, genotypes drawn at
the Table 2 frequencies, and Wuzhi capsule at the Table 1 rate (6/139). Each
child takes 1 mg every 12 h for 21 days; the last dosing interval is sampled.
The paper's VPC (Figure 3) is plotted against time since the first dose on a
mixed dosing history and is not replicated; the figure below shows the
model-predicted spread over one steady-state interval at a fixed dose.

```{r cohort}
rxode2::rxSetSeed(20221115)
set.seed(20221115)
n_sub <- 200
cohort <- tibble(
  id = seq_len(n_sub),
  AGE = pmin(pmax(exp(rnorm(n_sub, log(5.3), 0.55)), 1.1), 15.6),
  ctla4 = sample(c("GG", "GA", "AA"), n_sub, replace = TRUE, prob = c(2, 29, 108)),
  cyp3a5 = sample(c("*1/*1", "*1/*3", "*3/*3"), n_sub, replace = TRUE, prob = c(18, 58, 63)),
  CONMED_WUZHI = rbinom(n_sub, 1, 6 / 139)
) |>
  mutate(
    SNP_CTLA4_RS4553808_GA = as.numeric(ctla4 == "GA"),
    SNP_CTLA4_RS4553808_AA = as.numeric(ctla4 == "AA"),
    CYP3A5_EXPR = as.numeric(cyp3a5 != "*3/*3"),
    treatment = ifelse(CYP3A5_EXPR == 1, "CYP3A5 expresser", "CYP3A5 non-expresser")
  )

t_obs <- t_last + c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 10, 12)
ev_cohort <- bind_rows(
  tidyr::expand_grid(id = cohort$id, time = seq(0, by = tau, length.out = ndose)) |>
    mutate(amt = 1, evid = 1L, cmt = "depot"),
  tidyr::expand_grid(id = cohort$id, time = t_obs) |>
    mutate(amt = 0, evid = 0L, cmt = "central")
) |>
  left_join(
    select(cohort, id, AGE, SNP_CTLA4_RS4553808_GA, SNP_CTLA4_RS4553808_AA, CYP3A5_EXPR, CONMED_WUZHI, treatment),
    by = "id"
  ) |>
  arrange(id, time, desc(evid))

sim <- rxode2::rxSolve(mod, events = ev_cohort, keep = "treatment", maxsteps = 1e6) |>
  as.data.frame() |>
  mutate(tad = time - t_last)
stopifnot(!anyNA(sim$Cc))
```

```{r profile, fig.cap = "Simulated steady-state whole-blood tacrolimus over one 12 h interval on 1 mg q12h: median and 5th-95th percentiles of the individual predictions, by CYP3A5 status."}
sim |>
  group_by(treatment, tad) |>
  summarise(
    q05 = quantile(Cc, 0.05), q50 = median(Cc), q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, q50, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.2, colour = NA) +
  geom_line() +
  labs(x = "Time after dose (h)", y = "Tacrolimus (ng/mL)", colour = NULL, fill = NULL)
```

# PKNCA validation

The paper reports no NCA. Typical-value subjects (`zeroRe()`) of each CYP3A5
status (5.3 years, CTLA4 AA, no Wuzhi capsule) receive 1 mg every 12 h; PKNCA
computes Cmax, Cmin and AUC over the last interval, and the AUC must equal the
closed form `Dose / (CL/F)`.

```{r pknca}
nca_grp <- tibble(
  id = 1:2, CYP3A5_EXPR = c(1, 0),
  treatment = c("CYP3A5 expresser", "CYP3A5 non-expresser"),
  AGE = 5.3, SNP_CTLA4_RS4553808_GA = 0, SNP_CTLA4_RS4553808_AA = 1, CONMED_WUZHI = 0
)
ev_nca <- bind_rows(
  tidyr::expand_grid(id = nca_grp$id, time = seq(0, by = tau, length.out = ndose)) |>
    mutate(amt = 1, evid = 1L, cmt = "depot"),
  tidyr::expand_grid(id = nca_grp$id, time = t_last + c(0, 0.1, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 10, 12)) |>
    mutate(amt = 0, evid = 0L, cmt = "central")
) |>
  left_join(nca_grp, by = "id") |>
  arrange(id, time, desc(evid))

sim_nca <- rxode2::rxSolve(mod_typ, events = ev_nca, keep = "treatment") |>
  as.data.frame()

conc <- sim_nca |>
  filter(!is.na(Cc), time >= t_last) |>
  select(id, time, Cc, treatment)
doses_nca <- ev_nca |>
  filter(evid == 1L, time == t_last) |>
  select(id, time, amt, treatment)

o_conc <- PKNCAconc(conc, Cc ~ time | treatment + id)
o_dose <- PKNCAdose(doses_nca, amt ~ time | treatment + id)
intervals <- data.frame(start = t_last, end = t_last + tau, auclast = TRUE, cmax = TRUE, cmin = TRUE, tmax = TRUE)
nca_res <- pk.nca(PKNCAdata(o_conc, o_dose, intervals = intervals))

cl_nca <- sim_nca |>
  filter(time == t_last) |>
  distinct(treatment, cl)
nca_wide <- as.data.frame(nca_res) |>
  select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  left_join(cl_nca, by = "treatment") |>
  mutate(auc_closed = 1000 * 1 / cl)

nca_wide |>
  select(treatment, cmax, tmax, cmin, auclast, auc_closed) |>
  dplyr::rename(
    "Group" = treatment, "Cmax (ng/mL)" = cmax, "Tmax (h)" = tmax,
    "Cmin (ng/mL)" = cmin, "AUCtau (PKNCA, ng*h/mL)" = auclast,
    "Dose/(CL/F) (ng*h/mL)" = auc_closed
  ) |>
  knitr::kable(digits = 2, caption = "Typical-value steady-state NCA on 1 mg q12h.")

# Same typical parameters on both sides; the only difference is trapezoidal
# error over a dense 14-point interval, so a tight bound applies.
stopifnot(nrow(nca_wide) == 2, all(abs(nca_wide$auclast / nca_wide$auc_closed - 1) < 0.02))
```

# Assumptions and deviations

* **IIV scale.** Table 3 labels the random-effect rows "omega2 V (%)" and
  "omega2 CL (%)" and its footnote calls them variances; they are encoded as
  variances (1.13 and 0.13). The V/F variance is large (SD 1.06 on the log
  scale) and has 38% shrinkage, which the authors attribute to the trough-rich
  sampling. No covariance between the two random effects is reported, so none
  is included.
* **Residual error.** Table 3 gives sigma = 29.95% and the footnote calls it
  "square root of residual variability"; the bootstrap median 0.30 confirms the
  SD-on-the-fraction scale, so `propSd = 0.2995`.
* **CTLA4 reference group.** Eq. 2 and the coding note make GG the reference
  genotype, although only 2 of 139 patients are GG. The non-monotone GA and AA
  coefficients are encoded as published.
* **CYP3A5 coding.** The paper's indicator is 1 for CYP3A5\*3/\*3. The model
  uses the package's `CYP3A5_EXPR` (1 = expresser) and applies the published
  coefficient to `1 - CYP3A5_EXPR`, so the typical value 10.54 L/h refers to a
  CYP3A5 expresser.
* **Age reference.** The 5.3-year reference in Eq. 2 is the cohort median
  (Table 1), as the Methods covariate form `(COV/COVmedian)^theta` prescribes.
* **Fixed ka.** 4.48 1/h is taken from earlier paediatric tacrolimus models
  and was not estimated.
* **Wuzhi capsule.** Taken per record; the paper does not say whether it
  varied within a patient.
* **Dose units.** The paper reports doses in ug and concentrations in ug/L;
  the model takes doses in mg and returns `Cc` in ng/mL (= ug/L).
* **Cohort assumptions.** The age distribution of the virtual cohort is not
  reported beyond the median and range; a truncated log-normal with log-scale
  SD 0.55 is assumed. Genotypes are drawn independently at the Table 2
  frequencies.
* **Table 1 misprints.** The HCT range is printed as 23.4-447.6 %; HCT was not
  retained, so this does not affect the model. Gender is printed as 37/102
  (male/female) and is reported here as printed.
