---
title: "Tiapride (Huang 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tiapride (Huang 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r libraries, message = FALSE, warning = FALSE}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(tidyr)
library(ggplot2)

rxode2::rxSetSeed(20260913)

mod <- readModelDb("Huang_2026_tiapride")
ui <- rxode2::rxode(mod)
```

## The paper

Huang and colleagues developed a population pharmacokinetic model for
**tiapride**, a benzamide-class D2 antagonist used for tic disorders, in 38
Chinese children and adolescents aged 5-15 years followed as outpatients at a
single centre. Two models are reported and **both are implemented here, in one
model file**, because the authors built them as one nested structure rather
than as two competing alternatives: a plasma model (Table 2, Equations 5-7),
and then a plasma-saliva joint model (Table 3, Equations 8-10) fitted
*sequentially* -- the plasma parameters were held fixed and a saliva
compartment was added on top with the NONMEM ADVAN6 subroutine. Nothing in the
plasma layer moves when the saliva layer is present, so a single file
reproduces Table 2 and Table 3 simultaneously.

The clinical point of the paper is the saliva layer: saliva is a non-invasive
matrix for therapeutic drug monitoring in children, and tiapride concentrates
in it (mean saliva:plasma ratio 2.3). The transfer is **saturable** rather than
proportional -- a linear transfer model "resulted in marked overprediction of
saliva concentration in the high-concentration range" -- which is why the
saliva:plasma ratio is itself concentration-dependent and why saliva-based
monitoring needs a model rather than a multiplier.

One quantity the saliva layer needs is not printed anywhere in the paper: the
scale converting the saliva compartment amount to the reported ng/mL. It is
recovered here from the published visual predictive check, and it is the only
assumed value in the model file. **Read the Errata before using the saliva
output quantitatively.**

Reference:

```{r reference, echo = FALSE, results = "asis"}
cat(ui$reference, "\n")
```

## Population

```{r population, echo = FALSE}
pop <- ui$population
knitr::kable(
  data.frame(
    Field = names(pop),
    Value = vapply(pop, function(x) paste(
      if (is.null(names(x))) x else paste0(names(x), " ", x),
      collapse = "; "
    ), character(1)),
    row.names = NULL
  ),
  caption = "Study population (Huang 2026 Table 1 and Methods)."
)
```

Sampling was opportunistic and tied to clinic visits after at least seven days
of continuous treatment, so all 214 analysed plasma samples are at steady
state. Twenty-one percent fell in the absorption phase (0-2 h) and 49% beyond
10 h post-dose, leaving the 2.5-10 h window sparse -- which is why a
two-compartment model was unstable and the authors selected a one-compartment
structure despite the biphasic disposition reported for tiapride in adults.

## Source trace

Every equation and every `ini()` value, with the location it came from.
**Equations 1-10 of the source are vector graphics that both `pdftotext` and
the markdown preprocessor drop silently**; they were recovered by rendering
pages 3-8 of the PDF at 200-500 dpi and reading them as images.

```{r source-trace, echo = FALSE}
knitr::kable(
  tibble::tribble(
    ~Item,                       ~Value,                    ~Source,
    "Ka",                        "0.219 1/h",               "Table 2; Equation 5",
    "CL/F",                      "15.3 L/h",                "Table 2; Equation 6",
    "Vd/F",                      "5.77 L",                  "Table 2; Equation 7",
    "FFM exponent on CL/F",      "0.553",                   "Table 2; Equation 6",
    "FFM reference value",       "30.62 kg",                "Equation 6 denominator = Table 1 cohort median FFM",
    "IIV Ka",                    "17.6% -> omega = 0.176",  "Table 2 'eta Ka (%)'",
    "IIV CL/F",                  "22.8% -> omega = 0.228",  "Table 2 'eta CL/F (%)'",
    "IIV Vd/F",                  "84.3% -> omega = 0.843",  "Table 2 'eta Vd/F (%)'",
    "Proportional residual",     "15.6%",                   "Table 2 'eps prop (%)'",
    "Additive residual",         "87.9 ng/mL = 0.0879 mg/L","Table 2 'eps add (ng/mL)'",
    "Structural model",          "1-cmt, first-order absorption, linear elimination", "Results; Figure 2 (left half)",
    "IIV model",                 "exponential, P = theta * exp(eta)",  "Methods Equations 1-3",
    "Residual model",            "combined proportional + additive",   "Results 'a combined error model'",
    "Covariate model",           "power function on continuous covariates", "Methods Equation 3",
    "Therapeutic window",        "560-2000 ng/mL steady-state Cmax",   "Methods Model-Based Simulations (ref 13)",
    "PTA, 100 mg bid / 75 mg tid / 100 mg tid", "98.2% / 97.1% / 97.3%", "Results Monte Carlo Simulations"
  ),
  caption = "Source trace for the Huang 2026 plasma model (Table 2)."
)
```

The saliva layer, fitted sequentially on top of the fixed plasma parameters:

```{r source-trace-saliva, echo = FALSE}
knitr::kable(
  tibble::tribble(
    ~Item,                        ~Value,                       ~Source,
    "Vmax (central -> saliva)",   "34.7 mg/h",                  "Table 3; Equation 8",
    "Km (central -> saliva)",     "762 ng/mL = 0.762 mg/L, held constant", "Table 3; Equation 9",
    "K30 (saliva elimination)",   "6.24 1/h",                   "Table 3; Equation 10",
    "FFM exponent on K30",        "0.38",                       "Table 3; Equation 10",
    "IIV K30",                    "30% -> omega = 0.30",        "Table 3 'eta K30 (%)'",
    "Saliva proportional residual", "28.4%",                    "Table 3 'eps prop (%)'",
    "Transport structure",        "Michaelis-Menten, saturable, DRIVEN (no loss from central)", "Figure 2 (dashed arrow); Results; plasma parameters held fixed",
    "Saliva elimination",         "first-order out of saliva",  "Figure 2; Results",
    "Back-transfer saliva -> central", "none (tested, not retained)", "Discussion paragraph 6",
    "Apparent saliva volume",     "1.35 L -- NOT REPORTED, recovered from Figure 4", "See Errata; operator-ratified 2026-09-21",
    "Mean saliva:plasma ratio",   "2.3",                        "Discussion paragraph 2",
    "Saliva assay range",         "4-2000 ng/mL",               "Methods 'Sample analysis'"
  ),
  caption = "Source trace for the Huang 2026 saliva layer (Table 3)."
)
```

## Model

```{r model-print}
ui
```

### Flip-flop disposition

The estimated elimination rate constant is an order of magnitude larger than
the absorption rate constant, so the observed terminal slope reports
**absorption**, not elimination.

```{r flipflop}
theta <- setNames(ui$theta, names(ui$theta))
ka_tv <- exp(theta[["lka"]])
cl_tv <- exp(theta[["lcl"]])
vc_tv <- exp(theta[["lvc"]])
kel_tv <- cl_tv / vc_tv

c(ka = ka_tv, kel = kel_tv,
  t_half_absorption = log(2) / ka_tv,
  t_half_elimination = log(2) / kel_tv)
```

`ln(2)/Ka` = `r round(log(2) / ka_tv, 2)` h reproduces the terminal half-life
of 3.23 h that the paper quotes for tiapride in adults (Discussion, ref 35) --
the reconciliation the authors themselves make.

A guard that the `cl` / `vc` pair has not silently been replaced by rxode2's
analytic solution, discarding the explicit ODEs:

```{r lincmt-guard}
stopifnot(length(ui$linCmt) == 0L)
```

## Typical-value single dose: closed-form gates

The typical-value solve is deterministic, so it is checked against exact
closed-form identities rather than tolerance bands.

Note on `cmt` in the event tables below. This model declares **two**
endpoints (`Cc` and `Csaliva`), so rxode2 assigns each its own observation
slot and requires observation records to name the *observable*, not the ODE
state: `cmt = "central"` fails with `'dvid'->'cmt' or 'cmt' on observation
record or on a undefined compartment`. The usual library convention of
observing on the ODE state applies to single-endpoint models; here
`cmt = "Cc"` is required and correct. `useLinCmt = FALSE` is passed to every
`rxSolve()` for the same reason -- rxode2's automatic conversion to an
analytic solution corrupts the endpoint mapping for multi-output models. The
static vignette linter flags `cmt = "Cc"`; that warning is a known false
positive for multi-endpoint models and the render is the authority.

```{r typical-solve}
dose_mg <- 75

ev_typ <- data.frame(
  id   = 1L,
  time = c(0, sort(unique(c(seq(0, 4, by = 0.02), seq(0, 48, by = 0.1))))),
  amt  = NA_real_,
  evid = 0L,
  cmt  = "Cc",
  FFM  = 30.62
)
ev_typ$amt[1] <- dose_mg
ev_typ$evid[1] <- 1L
ev_typ$cmt[1] <- "depot"

sim_typ <- rxode2::rxSolve(rxode2::zeroRe(mod), events = ev_typ,
                           keep = "FFM", useLinCmt = FALSE) |>
  as.data.frame()
if (is.null(sim_typ$id)) sim_typ$id <- 1L

# Closed form for a one-compartment first-order-absorption model.
tmax_cf <- log(kel_tv / ka_tv) / (kel_tv - ka_tv)
cmax_cf <- dose_mg * ka_tv / (vc_tv * (ka_tv - kel_tv)) *
  (exp(-kel_tv * tmax_cf) - exp(-ka_tv * tmax_cf))
aucinf_cf <- dose_mg / cl_tv

c(tmax_closed_form = tmax_cf, cmax_closed_form = cmax_cf,
  aucinf_closed_form = aucinf_cf)
```

```{r typical-nca, warning = FALSE}
nca_conc <- sim_typ |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(regimen = "75 mg single dose") |>
  dplyr::select(id, time, Cc, regimen)

nca_conc <- dplyr::bind_rows(
  nca_conc,
  nca_conc |> dplyr::distinct(id, regimen) |>
    dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, regimen, time, .keep_all = TRUE) |>
  dplyr::arrange(id, regimen, time)

nca_dose <- data.frame(id = 1L, time = 0, amt = dose_mg,
                       regimen = "75 mg single dose")

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ time | regimen + id,
                             concu = "mg/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(nca_dose, amt ~ time | regimen + id,
                             doseu = "mg")

intervals_sd <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE,
  half.life = TRUE, lambda.z = TRUE
)

nca_sd <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                         intervals = intervals_sd))
nca_sd_tbl <- as.data.frame(nca_sd$result) |>
  dplyr::select(PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
nca_sd_tbl
```

```{r typical-gates}
stopifnot(
  # Mass balance: CL/F * AUC(0-inf) must return the administered dose exactly.
  # Catches a mis-transcribed clearance, a dropped unit conversion, and an
  # rxode2 auto-linCmt substitution that ignores the coded ODEs.
  abs(cl_tv * nca_sd_tbl$aucinf.obs / dose_mg - 1) < 0.005,
  # Terminal slope reports absorption (flip-flop), not elimination.
  abs(nca_sd_tbl$half.life - log(2) / ka_tv) < 0.05,
  # Cmax and Tmax against the exact closed form for this structure.
  abs(nca_sd_tbl$cmax / cmax_cf - 1) < 0.01,
  abs(nca_sd_tbl$tmax - tmax_cf) < 0.05,
  # The solve must not have drifted negative anywhere.
  all(sim_typ$Cc >= 0)
)
```

All four typical-value identities hold to better than 1%.

### Comparison against the values the paper quotes

```{r nca-comparison, warning = FALSE}
published <- tibble::tribble(
  ~regimen,             ~tmax, ~half.life,
  "75 mg single dose",  2.0,   3.23
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_sd,
  reference     = published,
  by            = "regimen",
  units         = c(tmax = "h", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated typical-value NCA against the tiapride descriptors Huang 2026",
    "quotes in its Discussion. * differs from reference by >20%."
  )
)
```

The terminal half-life matches to 2%. **Tmax is a starred deviation and is
expected to be**: 2 h is the observed peak time the paper reports for tiapride
(Discussion), while the fitted model puts the typical Tmax at
`r round(tmax_cf, 2)` h. The two are not the same quantity -- the observed 2 h
comes from the first VPC bin of an opportunistic outpatient sampling schedule
in which only 18 of 214 samples (8%) fall between 2 and 2.5 h, so the peak time
is poorly resolved by the data. No parameter was adjusted to close this gap.

## Virtual cohort and Monte Carlo simulation (replicates Figure 5)

The paper simulated 1000 virtual patients whose fat-free mass follows the
distribution of the modelled population, dosed for five consecutive days across
eight regimens, and read the peak concentration over the 12 h following the
last dose. FFM is drawn log-normally to reproduce the Table 1 median and IQR.

```{r cohort}
n_per_arm <- 200  # 200 per arm is the library cap; the paper used 1000

ffm_median <- 30.62
ffm_sdlog <- mean(c(log(34.65 / 30.62), log(30.62 / 26.93))) / qnorm(0.75)
c(ffm_sdlog = ffm_sdlog)

regimens <- tibble::tribble(
  ~label,          ~dose,  ~ii,
  "50 mg bid",     50.0,   12,
  "50 mg tid",     50.0,    8,
  "66.6 mg bid",   66.6,   12,
  "66.6 mg tid",   66.6,    8,
  "75 mg bid",     75.0,   12,
  "75 mg tid",     75.0,    8,
  "100 mg bid",   100.0,   12,
  "100 mg tid",   100.0,    8
)

make_arm <- function(label, dose, ii, id_offset) {
  subj <- data.frame(
    id  = id_offset + seq_len(n_per_arm),
    FFM = rlnorm(n_per_arm, meanlog = log(ffm_median), sdlog = ffm_sdlog)
  )
  doses <- tidyr::expand_grid(subj, time = seq(0, 120, by = ii)) |>
    dplyr::mutate(amt = dose, evid = 1L, cmt = "depot")
  obs <- tidyr::expand_grid(subj, time = seq(120, 132, by = 0.1)) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "Cc")
  dplyr::bind_rows(doses, obs) |>
    dplyr::mutate(regimen = label) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- do.call(
  dplyr::bind_rows,
  Map(make_arm, regimens$label, regimens$dose, regimens$ii,
      id_offset = (seq_len(nrow(regimens)) - 1L) * n_per_arm)
)

# Disjoint IDs across arms, so no cross-arm collision in the solve or in PKNCA.
stopifnot(
  dplyr::n_distinct(events$id) == nrow(regimens) * n_per_arm,
  events |> dplyr::distinct(id, regimen) |> nrow() ==
    nrow(regimens) * n_per_arm
)
```

```{r simulate}
sim <- rxode2::rxSolve(mod, events = events,
                       keep = c("regimen", "FFM"), useLinCmt = FALSE) |>
  as.data.frame()

sim_obs <- sim |> dplyr::filter(!is.na(Cc))
stopifnot(nrow(sim_obs) > 0, all(sim_obs$Cc >= 0))
```

`Cc` is the individual prediction and carries no residual error, so the peak
read off it is the model's Cmax rather than an upward-biased maximum of a
noisy series.

```{r figure5, fig.height = 6, fig.width = 8}
sim_obs <- sim_obs |>
  dplyr::mutate(
    Cc_ngml = Cc * 1000,
    regimen = factor(regimen, levels = regimens$label)
  )

ribbon <- sim_obs |>
  dplyr::group_by(regimen, time) |>
  dplyr::summarise(
    p05 = quantile(Cc_ngml, 0.05),
    p50 = median(Cc_ngml),
    p95 = quantile(Cc_ngml, 0.95),
    .groups = "drop"
  )

ggplot(ribbon, aes(time)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "#e8879b", alpha = 0.5) +
  geom_line(aes(y = p50), colour = "#b2182b", linewidth = 0.8) +
  geom_hline(yintercept = c(560, 2000), linetype = "dotted") +
  facet_wrap(~regimen, ncol = 4) +
  labs(
    x = "Time (h)", y = "Tiapride plasma concentration (ng/mL)",
    title = "Replicates Figure 5 of Huang 2026",
    subtitle = "Median and 5th-95th percentile over the 12 h after the last of five days of dosing"
  ) +
  theme_bw()
```

### Probability of target attainment

```{r pta}
cmax_ss <- sim_obs |>
  dplyr::group_by(regimen, id) |>
  dplyr::summarise(cmax_ngml = max(Cc_ngml), .groups = "drop")

pta <- cmax_ss |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(
    `Median Cmax (ng/mL)` = round(median(cmax_ngml)),
    `PTA, Cmax >= 560 ng/mL (%)` = round(100 * mean(cmax_ngml >= 560), 1),
    `Cmax > 2000 ng/mL (%)` = round(100 * mean(cmax_ngml > 2000), 1),
    .groups = "drop"
  )

knitr::kable(
  pta,
  caption = paste(
    "Simulated steady-state peak concentration by regimen.",
    "Huang 2026 reports PTA of 98.2%, 97.1% and 97.3% for 100 mg bid,",
    "75 mg tid and 100 mg tid respectively."
  )
)
```

```{r pta-gate}
pta_ref <- c(`100 mg bid` = 98.2, `75 mg tid` = 97.1, `100 mg tid` = 97.3)
pta_sim <- setNames(pta$`PTA, Cmax >= 560 ng/mL (%)`, pta$regimen)[names(pta_ref)]
stopifnot(!anyNA(pta_sim))  # a name mismatch must fail loudly, not pass vacuously

pta_diff <- pta_sim - pta_ref
pta_diff
```

```{r pta-assert}
stopifnot(
  # The three regimens the paper singles out must all clear 90% attainment.
  # A mis-transcribed clearance, dose or unit moves median Cmax by tens of
  # percent and drops these into the 60-80% range. The bound is deliberately
  # well outside the binomial noise of a 200-subject arm (SE ~1.2 points at
  # p = 0.97) plus the difference between this cohort's FFM draw and the
  # paper's own 1000-subject draw; do not tighten it to one observed run.
  all(pta_sim > 90),
  # 50 mg bid is the regimen the paper rejects as underdosing; it must sit
  # clearly below the three recommended ones.
  pta$`PTA, Cmax >= 560 ng/mL (%)`[pta$regimen == "50 mg bid"] <
    min(pta_sim) - 10,
  # Dose-ordering of the central tendency, which is structural.
  median(cmax_ss$cmax_ngml[cmax_ss$regimen == "100 mg tid"]) >
    median(cmax_ss$cmax_ngml[cmax_ss$regimen == "50 mg bid"])
)
```

The model reproduces the paper's central finding: 75 mg tid, 100 mg bid and
100 mg tid all attain the 560 ng/mL lower bound in nearly every subject, and
the two 100 mg regimens push a larger fraction above the 2000 ng/mL upper
bound -- which is exactly the efficacy/safety trade-off on which the authors
select 75 mg tid.

### Steady-state exposure gate

```{r ss-exposure}
# Over one dosing interval at steady state, CL/F * AUC(0-tau) = dose. Checked
# on the typical subject so the identity is exact rather than a cohort average.
ev_ss <- data.frame(
  id = 1L,
  time = c(seq(0, 120, by = 8),
           seq(120, 128, by = 0.02)),
  amt = c(rep(75, length(seq(0, 120, by = 8))),
          rep(NA_real_, length(seq(120, 128, by = 0.02)))),
  evid = c(rep(1L, length(seq(0, 120, by = 8))),
           rep(0L, length(seq(120, 128, by = 0.02)))),
  cmt = c(rep("depot", length(seq(0, 120, by = 8))),
          rep("Cc", length(seq(120, 128, by = 0.02)))),
  FFM = 30.62
) |>
  dplyr::arrange(time, dplyr::desc(evid))

sim_ss <- rxode2::rxSolve(rxode2::zeroRe(mod), events = ev_ss,
                          useLinCmt = FALSE) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc), time >= 120)

auc_tau <- sum(diff(sim_ss$time) *
                 (head(sim_ss$Cc, -1) + tail(sim_ss$Cc, -1)) / 2)

c(auc_tau = auc_tau, dose_over_cl = 75 / cl_tv,
  ratio = auc_tau / (75 / cl_tv))

stopifnot(abs(auc_tau / (75 / cl_tv) - 1) < 0.01)
```

At steady state the area over one 8 h interval returns `Dose / (CL/F)` to
better than 1%, confirming the dosing pathway and the FFM reference value are
wired correctly.

## The saliva layer

```{r saliva-params}
vmax_tv <- exp(theta[["lvmax_saliva"]])
km_tv   <- exp(theta[["lkm_saliva"]])
k30_tv  <- exp(theta[["lkel_saliva"]])
vsal_tv <- exp(theta[["lvsaliva"]])

c(Vmax_mg_per_h = vmax_tv, Km_ng_per_mL = 1000 * km_tv,
  K30_per_h = k30_tv, saliva_half_life_min = 60 * log(2) / k30_tv,
  vsaliva_L = vsal_tv)
```

### Saliva is driven, not mass-balance-coupled

The single most consequential structural decision in this layer is whether the
Michaelis-Menten transport **removes** drug from the central compartment. It
does not, and the model file encodes it that way. Three independent facts fix
the reading, and the third is checkable here:

1. Figure 2 draws this one arrow **dashed** where every other arrow in the
   diagram is solid.
2. The layer was fitted with the plasma parameters held **fixed**, so a term
   that drained central would have invalidated them -- yet the plasma
   goodness-of-fit in Figure 3A is unchanged from the plasma-only model.
3. The printed constants are **not mass-conserving**. `Vmax` = 34.7 mg/h
   exceeds the entire cohort-average absorption rate (about 9 mg/h at the
   median 215 mg/day dose), so a depleting transport term would remove more
   drug than is ever absorbed.

If the transport drained central, the plasma curve would depart from the
one-compartment closed form -- most at the peak, where the transport term is
largest. It does not:

```{r saliva-driven-gate}
cc_closed <- dose_mg * ka_tv / (vc_tv * (ka_tv - kel_tv)) *
  (exp(-kel_tv * sim_typ$time) - exp(-ka_tv * sim_typ$time))
max_rel_dev <- max(abs(sim_typ$Cc - cc_closed)) / max(cc_closed)

c(max_relative_deviation_from_closed_form = max_rel_dev)

stopifnot(
  # Exact identity, not a tolerance band: the plasma layer is numerically
  # untouched by the presence of the saliva compartment.
  max_rel_dev < 1e-4
)
```

Together with the `CL/F * AUC(0-inf) = dose` gate above -- which also holds
only if no mass leaves central -- this is the numerical proof of the
structural reading.

### Quasi-steady state

`K30` = `r round(k30_tv, 2)` 1/h is a
`r round(60 * log(2) / k30_tv, 1)`-minute half-life, so saliva equilibrates
with plasma far faster than plasma itself moves. Away from the steep
absorption phase the saliva state therefore tracks

$$A_{saliva} \approx \frac{V_{max} \cdot C_c}{(K_m + C_c) \cdot K_{30}}$$

```{r saliva-qss}
qss_amt <- vmax_tv * sim_typ$Cc / ((km_tv + sim_typ$Cc) * k30_tv)
qss_conc <- qss_amt / vsal_tv

# Restrict to the post-peak window: during the first minutes after a dose the
# plasma concentration is changing faster than saliva can follow, so the
# quasi-steady-state approximation is not expected to hold there.
post_peak <- sim_typ$time >= tmax_cf
qss_dev <- max(abs(sim_typ$Csaliva[post_peak] - qss_conc[post_peak])) /
  max(sim_typ$Csaliva)

c(max_relative_qss_deviation_after_tmax = qss_dev)

stopifnot(qss_dev < 0.05)
```

This matters because the quasi-steady-state form is the algebra used to
recover the unreported saliva volume (Errata below); the check confirms the
approximation is sound in the region where it was applied.

### The saliva:plasma ratio is concentration-dependent

This is the paper's central pharmacological claim. Because saliva is at quasi
steady state, the ratio has a closed form that is **independent of dose and of
time**:

$$\frac{C_{saliva}}{C_c} = \frac{V_{max}}{(K_m + C_c) \cdot K_{30} \cdot V_{saliva}}$$

It falls as plasma concentration rises -- the saturation the paper attributes
to a ceiling on salivary ion trapping.

```{r saliva-ratio}
sp_ratio <- function(cc_mg_per_L) {
  vmax_tv / ((km_tv + cc_mg_per_L) * k30_tv * vsal_tv)
}

# Huang 2026 Figure 1 fits two separate linear regressions to the paired
# samples, split at 700 ng/mL. Those give an entirely independent read of the
# same ratio -- neither regression was used to build this model.
fig1_low  <- function(cc_ng) (2.27 * cc_ng + 10.14) / cc_ng
fig1_high <- function(cc_ng) (1.04 * cc_ng + 885) / cc_ng

ratio_tbl <- tibble::tibble(
  plasma_ng_mL = c(200, 400, 700, 1000, 1400, 1700)
) |>
  dplyr::mutate(
    model  = sp_ratio(plasma_ng_mL / 1000),
    figure1 = ifelse(plasma_ng_mL < 700,
                     fig1_low(plasma_ng_mL), fig1_high(plasma_ng_mL))
  )

knitr::kable(
  ratio_tbl |>
    dplyr::rename(
      "Plasma (ng/mL)" = plasma_ng_mL,
      "Model S/P" = model,
      "Figure 1 regression S/P" = figure1
    ),
  digits = 2,
  caption = paste(
    "Model saliva:plasma ratio against the two paired-sample regressions of",
    "Huang 2026 Figure 1."
  )
)
```

```{r saliva-ratio-gates}
sp_at_peak <- sp_ratio(max(sim_typ$Cc))

c(model_SP_at_typical_peak = sp_at_peak,
  paper_mean_SP = 2.3,
  model_SP_at_1700 = sp_ratio(1.7),
  figure1_SP_at_1700 = fig1_high(1700))

stopifnot(
  # Saturation direction: the ratio must DECREASE with concentration. This is
  # the paper's qualitative central finding and is independent of the assumed
  # saliva volume, which scales the ratio but cannot change its shape.
  sp_ratio(0.4) > sp_ratio(1.6),
  # Saliva exceeds plasma everywhere in the observed range (S/P > 1), the
  # observation that motivates saliva monitoring at all.
  sp_ratio(1.7) > 1,
  # The mean saliva:plasma ratio of 2.3 that the Discussion reports for the
  # paired samples was NOT used to set the saliva volume (that anchor was
  # rejected -- see Errata), so this is an out-of-sample check. The model puts
  # the ratio at the typical peak within 20% of it.
  abs(sp_at_peak / 2.3 - 1) < 0.20,
  # At the top of the observed range the model tracks the Figure 1 regression
  # to within 25%.
  abs(sp_ratio(1.7) / fig1_high(1700) - 1) < 0.25
)
```

The agreement is good at high concentrations and deteriorates at low ones:
the model puts S/P near `r round(sp_ratio(0.2), 1)` at 200 ng/mL where the
Figure 1 regression implies about `r round(fig1_low(200), 1)`. That
discrepancy is **inherited from the published constants, not introduced
here** -- the shape of the S/P curve is fixed entirely by `Vmax`, `Km` and
`K30`, and the assumed saliva volume only slides the whole curve up or down.
With `Km` = 762 ng/mL sitting inside the observed plasma range, the model
predicts a several-fold swing in S/P across the data where the paired data
show a milder one. See the Errata.

### Plasma and saliva profiles (replicates Figure 4)

```{r saliva-profiles, fig.height = 4, fig.width = 7}
profile_tbl <- sim_typ |>
  dplyr::filter(time <= 12) |>
  dplyr::transmute(
    time,
    Plasma = 1000 * Cc,
    Saliva = 1000 * Csaliva
  ) |>
  tidyr::pivot_longer(c(Plasma, Saliva),
                      names_to = "Matrix", values_to = "conc")

ggplot2::ggplot(profile_tbl,
                ggplot2::aes(time, conc, colour = Matrix)) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::labs(
    x = "Time after dose (h)",
    y = "Tiapride concentration (ng/mL)",
    colour = NULL,
    title = "Typical-value profiles after a single 75 mg dose",
    subtitle = "Saliva exceeds plasma throughout, as in Huang 2026 Figure 4"
  ) +
  ggplot2::theme_bw()
```

Saliva peaks later and higher than plasma, and decays in parallel with it once
quasi steady state is reached -- the qualitative shape of Figure 4B.

```{r saliva-profile-gates}
sal_peak_time <- sim_typ$time[which.max(sim_typ$Csaliva)]

c(plasma_tmax = tmax_cf, saliva_tmax = sal_peak_time,
  saliva_peak_ng_mL = 1000 * max(sim_typ$Csaliva))

stopifnot(
  # Saliva lags plasma: the transport is driven by Cc and the saliva state
  # needs a few of its own half-lives to catch the rising plasma curve.
  sal_peak_time > tmax_cf,
  # but only slightly, given the 6.7-minute saliva half-life.
  sal_peak_time - tmax_cf < 0.5,
  # Saliva above plasma from the peak onwards. NOT asserted before the peak:
  # the saliva compartment starts empty and needs a few of its own half-lives
  # to fill, so Csaliva is legitimately below Cc during the first minutes
  # after the dose. The paper's paired samples were all drawn at clinic
  # visits well past that window.
  all(sim_typ$Csaliva[post_peak] >= sim_typ$Cc[post_peak]),
  all(sim_typ$Csaliva >= 0)
)
```

## Covariate effect of fat-free mass

```{r ffm-effect}
ffm_grid <- seq(18, 50, by = 1)
cl_by_ffm <- cl_tv * (ffm_grid / 30.62)^ui$theta[["e_ffm_cl"]]

ggplot(data.frame(FFM = ffm_grid, CL = cl_by_ffm), aes(FFM, CL)) +
  geom_line(linewidth = 0.8) +
  geom_point(data = data.frame(FFM = 30.62, CL = cl_tv), size = 2.5) +
  labs(x = "Fat-free mass (kg)", y = "Apparent clearance CL/F (L/h)",
       title = "Equation 6: CL/F = 15.3 * (FFM/30.62)^0.553") +
  theme_bw()

# Across the cohort IQR, clearance spans a factor of:
round((34.65 / 26.93)^ui$theta[["e_ffm_cl"]], 3)
```

Over the interquartile range of fat-free mass in the study population,
apparent clearance varies by only about 14%, consistent with the modest
exponent and its wide bootstrap confidence interval (0.277-0.771).

## Assumptions and deviations

### The saliva volume is assumed, not reported

**This is the one assumed value in the model file, and it sits directly under
the saliva layer's primary output. Read this section before using `Csaliva`
quantitatively.**

Every other saliva parameter is printed: `Vmax` = 34.7 mg/h (Equation 8),
`Km` = 762 ng/mL held constant (Equation 9), and
`K30 = 6.24 * (FFM/30.62)^0.38 * exp(eta)` (Equation 10), with 30% IIV on K30
and 28.4% proportional residual error (Table 3). Those three constants fully
determine the *amount* of drug in the saliva compartment. Converting that
amount to the observed saliva *concentration* needs one more quantity -- an
apparent saliva volume -- and it appears nowhere in the paper: not in
Tables 1-3, not in Equations 1-10 (all ten were rendered from the PDF's vector
graphics and read), not in any figure caption, and not in the Methods. In
NONMEM this would have been a `$SCALE`/`S3` statement rather than a `$THETA`,
which is very likely why it never reached the parameter table.

There is nothing to acquire: the EuropePMC `supplementaryFiles` endpoint
returns HTTP 404 for `PMC13111164`, the core record reports `hasSuppl: N`, and
the article carries no data-availability or supplementary-material statement.
This is a reporting gap, not an acquisition gap.

Writing the quasi-steady-state form
`Csaliva = Vmax * Cc / ((Km + Cc) * K30 * Vsaliva)`, `Vsaliva` is the only
unknown, and candidate recoveries span more than threefold depending on which
published quantity is used to pin it:

| Basis for recovery | Implied `Vsaliva` | Disposition |
|---|---|---|
| Model median saliva / plasma at the Figure 4 VPC peak | **1.35 L** | **Adopted** |
| NONMEM's default `S3 = 1` (a unit-volume compartment) | 1 L | Rejected -- puts the low-concentration S/P near 7.3 against a reported mean of 2.3, the worst agreement with the paper's own data of any candidate |
| Fitting the Figure 1 regressions to the saturating form | ~1.9 L | Rejected -- the two regressions are not mutually consistent with a single saturating curve |
| The reported mean saliva:plasma ratio of 2.3 | ~3.2 L | Rejected -- exact only in the low-concentration limit, and puts saliva about 35% below the Figure 4B median at the peak |
| Sharing the central volume, as `Nguyen_2026_linezolid` does | 5.77 L | Falsified -- gives saliva *below* plasma, contradicting the paper's central finding |

**The adopted value and its provenance.** `Vsaliva` = 1.35 L was recovered by
digitising Figure 4 at 500 dpi. At the roughly 2 h peak the model's median
plasma prediction is about 790 ng/mL and its median saliva prediction about
2100 ng/mL; substituting into the quasi-steady-state form gives
`A_saliva` = 34.7 x 0.790 / ((0.762 + 0.790) x 6.24) = 2.83 mg and hence
`Vsaliva` = 1000 x 2.83 / 2100 = 1.35 L. The value is annotated inline in
`ini()` as figure-derived. This follows the standing figure-digitisation
policy for parameters available only graphically, and was
**operator-ratified on 2026-09-21** (sidecar `oasweep_PMC13111164`
`request-001` / `response-001`, question q1, option B).

**What this costs.** Three caveats, in decreasing order of importance:

1. *The Figure 4 peak is calibration, not validation.* `Vsaliva` was chosen to
   make the model reproduce that point, so agreement there is circular and is
   not claimed as a check anywhere in this vignette. The genuinely independent
   comparisons are the mean saliva:plasma ratio of 2.3 from the Discussion and
   the Figure 1 paired regressions, neither of which was used in the recovery;
   both are gated above.
2. *The recovery is uncertain by roughly twofold.* Applying the same algebra at
   trough times rather than at the peak yields about 2.4 L. The digitisation
   itself is good to perhaps +/- 25%, but the choice of anchor point matters
   more than the reading precision.
3. *`Vsaliva` cannot fix the shape.* It scales the whole S/P curve uniformly,
   so it slides agreement between low and high concentrations without ever
   achieving both. The model's S/P falls about threefold from 200 to 1700
   ng/mL, whereas the Figure 1 paired regressions imply a milder decline. That
   is a property of the published `Km` = 762 ng/mL sitting inside the observed
   plasma range, and it would be present at any `Vsaliva`.

Anyone needing quantitatively reliable saliva predictions should treat
`Csaliva` as accurate to within a factor of about two, use it in the
high-concentration region where it agrees best with Figure 1, and ideally
write to the corresponding authors (Xiaorong Yin, `sunnyxrr@126.com`; Xuemei
Wu, `wuxuemei@fjmu.edu.cn`) for the scale statement. The plasma layer carries
none of this uncertainty.

### Two saliva values differ between Table 3 and the Discussion

The Discussion states "the estimated K30 was 6.19 h-1" and "an IIV of 31.3% in
K30", where Table 3 and Equation 10 both print 6.24 1/h and 30%. The model
follows **Table 3 and Equation 10**, on the standing rule that a printed
equation and the final parameter table outrank a Discussion recollection. The
difference is well inside the bootstrap interval (5.72-6.8 for K30) and moves
no conclusion.

### The saliva transport is encoded as driven, not depleting

The paper never writes the central compartment's differential equation, so
whether the Michaelis-Menten term is subtracted from central is a
reconstruction. It is encoded as **driven** (central is untouched) on the
three grounds set out in the saliva section above -- the dashed arrow in
Figure 2, the sequential fit with plasma parameters held fixed, and the fact
that the printed `Vmax` is too large to be mass-conserving. The same structural
reading was operator-ratified for `Nguyen_2026_linezolid.R`, the other saliva
model in this library. Were the term instead depleting, the plasma layer would
no longer reproduce Table 2 and the `CL/F * AUC = dose` identity would fail by
tens of percent, so this is not a fine distinction.

### Much of the observed saliva data sits above the highest calibrator

The LC-MS/MS saliva assay was calibrated over 4-2000 ng/mL (Methods), but
Figure 4B shows observed saliva concentrations reaching about 3900 ng/mL and a
median above 2000 ng/mL at the peak. A substantial part of the saliva data the
model was fitted to therefore lies above the calibrated range. The paper does
not comment on this. It is a property of the source, not of the
implementation, but it bears on how much weight the saliva layer will carry.

### Interindividual variability scale

Table 2 heads the variability rows `eta Ka (%)`, `eta CL/F (%)` and
`eta Vd/F (%)` -- the eta itself as a percentage -- so they are read as
`omega * 100` and encoded as `omega^2`. The paper prints no CV% column and no
control stream that would settle the convention independently. Read instead as
a log-normal CV%, the two small terms barely move (17.6% -> 0.1745, 22.8% ->
0.2249) but Vd/F would fall from 0.843 to 0.7444. Given the 56% shrinkage on
`eta Vd/F` and the bootstrap interval spanning 30-133%, that term carries
little information either way.

### Other assumptions

* **Fat-free mass equation.** The paper does not state how the FFM column was
  derived. For a 5-15 year old cohort the Al-Sallami paediatric correction to
  the Janmahasatian adult formula is the usual choice, and a user supplying
  `FFM` should compute it that way. The virtual cohort here draws FFM directly
  from a log-normal matched to the Table 1 median and IQR, as the paper did.
* **Bioavailability.** The study is oral-only, so `CL/F` and `Vd/F` are
  apparent. No `F` term is reported and none is encoded.
* **Units.** The model works in mg / L / h, so `Cc` is in mg/L; the paper
  reports ng/mL throughout and 1 mg/L = 1000 ng/mL. The additive residual
  87.9 ng/mL is encoded as 0.0879 mg/L.
* **Dosing schedule for Figure 5.** The paper says "simulated for 5 consecutive
  days" and plots 120-132 h, so the last dose is at t = 120 h with observation
  over the following 12 h and no further doses; that is what is reproduced.
* **PTA definition.** The paper's phrasing ("the proportion of simulated
  steady-state peak concentration within the therapeutic window of
  560-2000 ng/mL") is ambiguous, but only the lower-bound reading
  `P(Cmax >= 560)` reproduces the reported 97-98% values while also leaving
  room for the paper's separate observation that the 100 mg regimens push more
  patients above 2000 ng/mL. Both quantities are tabulated above.
* **Internal inconsistencies in the source.** The Discussion states
  `K30 = 6.19 1/h` and an IIV of 31.3% where Table 3 gives 6.24 and 30%, and
  says the optimal regimen is "75 mg bid" where the Abstract, Results and
  Conclusions all say 75 mg tid. The tables and the majority text are taken as
  authoritative. Table 1 prints BSA in "cm2" and BMI in "kg cm2", both of which
  are typographical errors for m^2 and kg/m^2.

## Session information

```{r session, echo = FALSE}
sessionInfo()
```
