---
title: "Colistin sulfate (Huang 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Colistin sulfate (Huang 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Huang_2025_colistinSulfate"))$reference`
* Description: `r rxode2::rxode(readModelDb("Huang_2025_colistinSulfate"))$description`
* Article: <https://doi.org/10.1177/00368504251325334>
* Supplement (Table S1, Figures S1-S2) and the publisher's native equation /
  table renderings: EuropePMC
  <https://www.ebi.ac.uk/europepmc/webservices/rest/PMC11877486/supplementaryFiles>

Huang 2025 is, by the authors' account, the first population PK analysis of
**colistin sulfate** in patients receiving continuous renal replacement therapy
(CRRT). Colistin sulfate is the *active* drug; it must not be confused with
colistimethate sodium (CMS), the inactive prodrug modelled elsewhere in this
package (`Plachouras_2009_colistin`, `Mohamed_2012_colistin`,
`Jacobs_2016_colistin`, `Karaiskos_2015_colistin`). The sibling model
`Ma_2026_colistinSulfate` describes the same active drug in an ICU population
from which CRRT and ECMO patients were *excluded*, so the two are complementary
rather than competing.

Dose unit conversion used throughout: **1 million unit (MU) = 44 mg**
(Methods, and the Figure 5 note).

# Population

Twenty critically ill adults at a single centre in Nanning, Guangxi, China
(May 2023 - January 2024; ChiCTR2300072191) contributed 86 plasma
concentrations. All were receiving CVVHDF for acute kidney injury for at least
48 h and had confirmed or suspected carbapenem-resistant organism (CRO)
infection. The cohort was severely ill: SOFA 9.27 +/- 4.08, APACHE II
26.6 +/- 10.1, with 90% on vasoactive agents, 90% mechanically ventilated, 90%
with lung infection and 75% with multi-site infection (Table 1). Age was
50.5 +/- 14.1 years, weight 68.6 +/- 14.1 kg (median 65 kg) and 75% were male.
Cystatin C was 2.87 +/- 1.91 mg/L (median 2.31 mg/L, observed range
1.05-5.11 mg/L per the Figure 5 note).

Sampling was rich *within* a maintenance-phase dosing interval: a pre-dose
trough plus four to six post-dose samples at nominally 0.5, 1, 2, 4 and 6 h.
Most subjects (85%) received 1.5 MU daily divided q8h, infused over 0.5 h.

The same information is available programmatically via
`readModelDb("Huang_2025_colistinSulfate")()$population`.

```{r population}
pop <- rxode2::rxode(readModelDb("Huang_2025_colistinSulfate"))$population
str(pop, max.level = 1, give.attr = FALSE)
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Huang_2025_colistinSulfate.R` carries an in-file
comment naming its origin. Collected here for review:

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL)           | 3.69 L/h | Table 3, "Final model / Estimate" (RSE 22.2%); also the Results final-model equation |
| `lvc` (V1)           | 20.50 L | Table 3 (RSE 10.1%); Results equation |
| `lvp` (V2)           | 33.20 L | Table 3 (RSE 17.3%); Results equation |
| `lq`  (Q)            | 25.30 L/h | Table 3 (RSE 26.5%); Results equation |
| `e_cysc_cl`          | -0.626 | Table 3 `theta_CysC` (RSE 16.8%) |
| `e_wt_cl`            | 1.03 | Table 3 `theta_wt` (RSE 21.7%) |
| CYSC centring        | 2.31 mg/L | Results text: "The values 2.31 and 65 were the medians for CYSC and WT" |
| WT centring          | 65 kg | Results text, same sentence |
| `etalcl`             | 0.31^2 | Table 3 `IIV_CL` 31% as an omega SD (bootstrap median 0.311) |
| `etalvp`             | 0.545^2 | Table 3 `IIV_V2` 54.5% as an omega SD (bootstrap median 0.557) |
| `expSd`              | 0.203 | Table 3 `RSV` 20.3% as a log-scale SD (bootstrap median 0.193) |
| Two-compartment, first-order linear elimination | n/a | Results "Population PK model"; supplement Table S1 "Structure model" |
| IIV on CL and V2 only | n/a | Results (IIV on V1 and Q near zero, shrinkage > 30%); supplement Table S1 |
| Exponential residual error | n/a | Methods; Results; supplement Table S1 ("RSV (exponential model)") |
| CL covariate equation | see below | Results final-model equation; publisher rendering `...-eq2.jpg` |

The final-model equation, transcribed from the publisher's own rendering of the
Results equation block:

$$
\mathrm{CL\ (L/h)} = 3.69 \times \left(\frac{\mathrm{CYSC}}{2.31}\right)^{-0.626}
\times \left(\frac{\mathrm{WT}}{65}\right)^{1.03}
$$

**Both covariates are power terms.** The Results narrative says "CYSC and WT
are included on CL in power form and exponential form, respectively", but the
paper's own printed equation shows WT as a power term, and an exponential
reading is arithmetically impossible (see *Assumptions and deviations*).

# Structural checks

These gates are deterministic - they compare the packaged model against the
paper's printed equation, transcribed here independently. They do not depend on
any random draw, so they are asserted tightly.

```{r structural-checks}
mod <- readModelDb("Huang_2025_colistinSulfate")

# The paper's printed equation, transcribed directly from the Results block.
cl_paper <- function(cysc, wt) 3.69 * (cysc / 2.31)^-0.626 * (wt / 65)^1.03

# Gate 1: at the reference covariates the equation must return the typical
# value the paper tabulates in Table 3.
stopifnot(isTRUE(all.equal(cl_paper(2.31, 65), 3.69, tolerance = 1e-12)))

# Gate 2: the packaged model must reproduce that equation everywhere in the
# covariate space the paper simulated over (CysC 1.05-5.11, WT 50-80). Solve
# the typical-value model on a covariate grid and read back its own `cl`.
grid <- tidyr::expand_grid(
  CYSC = c(1.05, 1.40, 2.31, 3.07, 5.11),
  WT   = c(50, 65, 80)
) |>
  dplyr::mutate(id = dplyr::row_number())

grid_ev <- grid |>
  tidyr::crossing(time = c(0, 1)) |>
  dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central") |>
  dplyr::arrange(id, time)

grid_sim <- rxode2::rxSolve(
  rxode2::zeroRe(mod), events = grid_ev,
  keep = c("CYSC", "WT"), returnType = "data.frame"
)

chk_cl <- grid_sim |>
  dplyr::group_by(CYSC, WT) |>
  dplyr::summarise(cl_model = mean(cl), .groups = "drop") |>
  dplyr::mutate(
    cl_ref   = cl_paper(CYSC, WT),
    rel_diff = 100 * (cl_model - cl_ref) / cl_ref
  )

stopifnot(nrow(chk_cl) == 15L)                       # the gate had rows to test
stopifnot(max(abs(chk_cl$rel_diff)) < 1e-8)          # exact algebraic agreement

chk_cl |>
  dplyr::mutate(dplyr::across(c(cl_model, cl_ref), \(x) round(x, 4)),
                rel_diff = signif(rel_diff, 2)) |>
  dplyr::rename(
    "CysC (mg/L)"      = CYSC,
    "WT (kg)"          = WT,
    "CL, model (L/h)"  = cl_model,
    "CL, paper (L/h)"  = cl_ref,
    "% difference"     = rel_diff
  ) |>
  knitr::kable(caption = paste(
    "Packaged model vs the paper's printed CL equation over the simulated",
    "covariate space. Agreement is algebraically exact."
  ))
```

The covariate directions match the paper's prose ("clearance decreases as the
CysC level increases ... clearance increases with increased body WT"):
clearance spans `r sprintf("%.2f", min(chk_cl$cl_ref))` to
`r sprintf("%.2f", max(chk_cl$cl_ref))` L/h across the grid, a
`r sprintf("%.1f", max(chk_cl$cl_ref) / min(chk_cl$cl_ref))`-fold range.

```{r structural-derived}
# Derived disposition quantities implied by Table 3 (reference covariates).
vss   <- 20.5 + 33.2
# Effective (Vss-based) half-life; the true terminal slope is steeper.
t_half_eff <- log(2) * vss / 3.69
tibble::tibble(
  Quantity = c("Vss = V1 + V2 (L)", "Vss/WT at 65 kg (L/kg)",
               "Effective half-life log(2)*Vss/CL (h)"),
  Value = c(round(vss, 1), round(vss / 65, 3), round(t_half_eff, 1))
) |>
  knitr::kable(caption = "Disposition quantities derived from Table 3.")
```

A 0.78 L/kg steady-state volume is consistent with the paper's Discussion
remark that colistin sulfate has "a lower apparent distribution volume
[indicating] that the drug predominantly remains within the bloodstream".

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the Table 1 covariate distributions and the modal regimen (1.5 MU daily = 66 mg
daily, divided q8h as a 0.5-h infusion, which 85% and 90% of subjects
respectively received).

```{r cohort}
# `set.seed()` seeds R's RNG, NOT rxode2's simulation RNG, and rxode2's streams
# are partitioned per solver thread -- so this cohort is reproducible on this
# machine and different on a machine with a different thread count. Every
# assertion below is therefore written to hold for ANY cohort the model can
# produce (see pattern 12 of known-vignette-failure-patterns.md).
set.seed(20250829)

n_sub <- 200L   # <= 200 per arm (vignette cap)
tau   <- 8      # dosing interval (h)
dose  <- 22     # 0.5 MU = 22 mg per dose -> 66 mg (1.5 MU) daily
t_end <- 96     # >= 9 effective half-lives, so [72, 96] is steady state

# Table 1: WT 68.6 +/- 14.1 kg; CysC 2.87 +/- 1.91 mg/L. Cystatin C is
# strictly positive and right-skewed (mean 2.87 > median 2.31), so it is drawn
# log-normally with the moments matched to Table 1 and truncated to the
# observed 1.05-5.11 mg/L range quoted in the Figure 5 note.
cysc_sd_log <- sqrt(log(1 + (1.91 / 2.87)^2))
cysc_mu_log <- log(2.87) - cysc_sd_log^2 / 2

subj <- tibble::tibble(
  id   = seq_len(n_sub),
  WT   = pmin(pmax(rnorm(n_sub, 68.6, 14.1), 40), 110),
  CYSC = pmin(pmax(rlnorm(n_sub, cysc_mu_log, cysc_sd_log), 1.05), 5.11)
)

# Observation grid: coarse over the approach to steady state, dense over the
# [72, 96] window where NCA is computed so that Cmax is resolved.
obs_times <- sort(unique(c(seq(0, t_end, by = 0.25), seq(72, t_end, by = 0.05))))

dose_rows <- subj |>
  tidyr::crossing(time = seq(0, t_end - tau, by = tau)) |>
  dplyr::mutate(amt = dose, evid = 1L, dur = 0.5, cmt = "central")

obs_rows <- subj |>
  tidyr::crossing(time = obs_times) |>
  dplyr::mutate(amt = NA_real_, evid = 0L, dur = NA_real_, cmt = "central")

events <- dplyr::bind_rows(dose_rows, obs_rows) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod, events = events, keep = c("WT", "CYSC"), returnType = "data.frame"
)
if (is.null(sim$id)) sim$id <- 1L
stopifnot(all(sim$Cc[!is.na(sim$Cc)] >= 0))
```

## Replicates Figure 1 of Huang 2025

Figure 1 of the paper shows the observed concentration-time curves for the 16
subjects with sufficient sampling. The panel below is the simulated equivalent
over one steady-state day, with the observed cohort summaries overlaid as
reference lines.

```{r figure-1}
sim_ss <- sim |>
  dplyr::filter(time >= 72, time <= 96) |>
  dplyr::mutate(t_ss = time - 72)

sim_ss |>
  dplyr::group_by(t_ss) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05), Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(t_ss, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(linewidth = 0.8, colour = "steelblue4") +
  geom_hline(yintercept = 0.30, linetype = "dashed", colour = "grey30") +
  annotate("text", x = 0.4, y = 0.30, vjust = -0.6, hjust = 0, size = 3,
           colour = "grey30", label = "observed Css,min 0.30 mg/L") +
  labs(
    x = "Time within the steady-state day (h)", y = "Colistin (mg/L)",
    title = "Simulated steady-state profile, 0.5 MU q8h (1.5 MU daily)",
    subtitle = "Median with 5th-95th percentile band, n = 200 virtual subjects",
    caption = paste(
      "Compare Figure 1 of Huang 2025 (observed profiles, 16 subjects).",
      "Dashed line is the paper's observed mean Css,min."
    )
  )
```

## Clearance covariate effects

```{r figure-covariates}
sim |>
  dplyr::distinct(id, WT, CYSC, cl) |>
  ggplot(aes(CYSC, cl, colour = WT)) +
  geom_point(alpha = 0.8) +
  scale_colour_viridis_c(name = "WT (kg)") +
  labs(
    x = "Cystatin C (mg/L)", y = "Individual CL (L/h)",
    title = "Clearance falls with cystatin C and rises with body weight",
    caption = paste(
      "Reproduces the direction of both covariate effects reported in the",
      "Huang 2025 Results (theta_CysC = -0.626, theta_wt = +1.03)."
    )
  )
```

# PKNCA validation

Steady-state NCA over the last full day of dosing, `r 24 / 8`-dose interval
window [72, 96] h (PKNCA recipe 3).

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(time >= 72, time <= 96, !is.na(Cc)) |>
  dplyr::mutate(regimen = "0.5 MU q8h") |>
  dplyr::select(id, time, Cc, regimen)

conc_obj <- PKNCA::PKNCAconc(
  sim_nca, Cc ~ time | regimen + id, concu = "mg/L", timeu = "h"
)

dose_df <- events |>
  dplyr::filter(evid == 1, time >= 72, time < 96) |>
  dplyr::mutate(regimen = "0.5 MU q8h") |>
  dplyr::select(id, time, amt, regimen)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | regimen + id, doseu = "mg")

# One 24-h steady-state interval, matching the paper's AUCss,0-24h.
intervals <- data.frame(
  start = 72, end = 96,
  cmax = TRUE, cmin = TRUE, tmax = TRUE, auclast = TRUE, cav = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)
```

## Gate: NCA against the exact mass-balance identity

Integrating the ODE system over any window `[T1, T2]` gives an identity that
holds **exactly**, with no steady-state assumption:

$$\int_{T_1}^{T_2} C\,\mathrm{d}t \;=\; \frac{D_\mathrm{in} - \left[A(T_2) - A(T_1)\right]}{CL}$$

where `A = central + peripheral1` is the total amount in the body and
`D_in` is the dose administered during the window. Both compartment amounts
are returned by `rxSolve`, so this compares the PKNCA integral against the
solver's own mass balance using each subject's own clearance.

The steady-state form of this identity (`AUC = D_in / CL`) is **not** used as
the gate, because it is not true for every subject here: IIV on V2 is large
(omega = 0.545), so a subject drawing a big peripheral volume has a long
terminal half-life and has not fully accumulated by 72 h. The exact form above
is insensitive to that, which is why it is the gate.

```{r gate-massbalance}
per_id <- sim |>
  dplyr::group_by(id) |>
  dplyr::summarise(cl = dplyr::first(cl), .groups = "drop")

# Total amount in the body at each window edge. All three infusions in the
# window start at 72, 80 and 88 h and finish by 88.5 h, so D_in is exactly
# 3 x 22 mg and neither edge falls inside an infusion.
amt_edge <- sim |>
  dplyr::filter(time %in% c(72, 96)) |>
  dplyr::mutate(A = central + peripheral1) |>
  dplyr::select(id, time, A) |>
  tidyr::pivot_wider(names_from = time, values_from = A,
                     names_prefix = "A_")

auc_nca <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(id, auclast = PPORRES) |>
  dplyr::mutate(id = as.integer(as.character(id)))

chk_mb <- auc_nca |>
  dplyr::inner_join(per_id, by = "id") |>
  dplyr::inner_join(amt_edge, by = "id") |>
  dplyr::mutate(
    auc_exact = (3 * dose - (A_96 - A_72)) / cl,
    pct_diff  = 100 * (auclast - auc_exact) / auc_exact,
    # The steady-state approximation, shown for contrast only -- NOT gated.
    auc_ss    = 3 * dose / cl,
    pct_ss    = 100 * (auclast - auc_ss) / auc_ss
  )

stopifnot(nrow(chk_mb) == n_sub)          # the gate had rows to test
# Only trapezoidal error on a 0.05-h grid separates the two sides, and that is
# deterministic rather than a property of the drawn cohort. Realised max
# |diff| ~0.02%; 1% still catches a mis-specified dose, window or clearance,
# which move this by tens of percent.
stopifnot(max(abs(chk_mb$pct_diff)) < 1)

tibble::tibble(
  Quantity = c(
    "median PKNCA AUC0-24 over [72, 96] h (mg*h/L)",
    "median exact mass-balance AUC (mg*h/L)",
    "max |% difference|, exact identity (gated)",
    "max |% difference|, steady-state approximation (not gated)",
    "subjects >2% from the steady-state approximation"
  ),
  Value = c(
    round(median(chk_mb$auclast), 3),
    round(median(chk_mb$auc_exact), 3),
    signif(max(abs(chk_mb$pct_diff)), 2),
    signif(max(abs(chk_mb$pct_ss)), 2),
    sum(abs(chk_mb$pct_ss) > 2)
  )
) |>
  knitr::kable(caption = paste(
    "PKNCA AUC against the exact mass-balance identity (gated) and against",
    "the steady-state approximation (shown for contrast)."
  ))
```

The gap between the two right-hand columns is the point: the exact identity
agrees to a few hundredths of a percent for every subject, while the
steady-state approximation is off by double digits for the minority of
subjects still accumulating at 72 h.

## Comparison against published NCA

The paper reports two observed exposure summaries for the cohort
(Results, "Patient information and treatment outcomes"): Css,min
0.30 +/- 0.22 mg/L and AUCss,0-24h 12.51 +/- 6.41 mg*h/L, the latter obtained
for only 16 of 20 subjects by the trapezoidal rule.

```{r nca_compare}
published <- tibble::tibble(
  regimen = "0.5 MU q8h",
  cmin    = 0.30,    # Results: Css,min 0.30 +/- 0.22 mg/L
  auclast = 12.51    # Results: AUCss,0-24h 12.51 +/- 6.41 mg*h/L
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference  = published,
  by         = "regimen",
  params     = c("cmin", "auclast"),
  units      = c(cmin = "mg/L", auclast = "mg*h/L"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated (median of 200 virtual subjects) vs the paper's observed",
    "cohort summaries. * differs from the reference by >20%. Both rows are",
    "expected to be starred -- see Assumptions and deviations."
  )
)
```

Both rows differ from the observed summaries by more than 20%, in the same
direction: the packaged model predicts **higher** exposure than the paper's
observed means. This is a property of the published model itself, not of the
transcription, and it is quantified in the next chunk rather than tuned away.

```{r deviation-quantified}
# The paper's own parameters are internally consistent: at the cohort-mean
# covariates the printed equation gives CL = 3.41 L/h, so a 66 mg daily dose
# must produce AUCss,0-24h = 66 / 3.41 = 19.4 mg*h/L, not the 12.51 reported.
cl_cohort_mean <- cl_paper(2.87, 68.6)
auc_implied    <- 3 * dose / cl_cohort_mean

# Part of the gap is the paper's SAMPLING WINDOW. AUC was computed by the
# trapezoidal rule from samples at 0 (trough), 0.5, 1, 2, 4 and 6 h -- but the
# dosing interval is 8 h, so the 6-8 h tail is never measured. Quantify that
# truncation on the typical-value profile.
typ_ev <- tibble::tibble(id = 1L, CYSC = 2.87, WT = 68.6) |>
  (\(d) dplyr::bind_rows(
    d |> tidyr::crossing(time = seq(0, t_end - tau, by = tau)) |>
      dplyr::mutate(amt = dose, evid = 1L, dur = 0.5, cmt = "central"),
    d |> tidyr::crossing(time = sort(unique(c(seq(0, t_end, by = 0.25),
                                              seq(72, t_end, by = 0.01))))) |>
      dplyr::mutate(amt = NA_real_, evid = 0L, dur = NA_real_, cmt = "central")
  ))() |>
  dplyr::arrange(id, time, dplyr::desc(evid))

typ <- rxode2::rxSolve(rxode2::zeroRe(mod), events = typ_ev,
                       returnType = "data.frame") |>
  dplyr::filter(time >= 88, time <= 96) |>
  dplyr::mutate(t = time - 88)

trap <- function(t, c) sum(diff(t) * (head(c, -1) + tail(c, -1)) / 2)
auc_full_tau <- trap(typ$t, typ$Cc)
sample_grid  <- c(0, 0.5, 1, 2, 4, 6)
auc_trunc    <- trap(sample_grid, approx(typ$t, typ$Cc, sample_grid)$y)

tibble::tibble(
  Quantity = c(
    "CL at cohort-mean covariates (L/h)",
    "AUCss,0-24h implied by the paper's own CL (mg*h/L)",
    "AUCss,0-24h reported as observed (mg*h/L)",
    "observed / model-implied ratio",
    "true AUC over one 8-h interval (mg*h/L)",
    "trapezoid on the paper's 0-6 h sample grid (mg*h/L)",
    "truncation factor from stopping at 6 h"
  ),
  Value = c(
    round(cl_cohort_mean, 3), round(auc_implied, 2), 12.51,
    round(12.51 / auc_implied, 2), round(auc_full_tau, 3),
    round(auc_trunc, 3), round(auc_trunc / auc_full_tau, 2)
  )
) |>
  knitr::kable(caption = paste(
    "The published model over-predicts the paper's own observed AUC.",
    "Sampling truncation at 6 h in an 8-h interval accounts for part of it."
  ))

# Gate: the packaged model must agree with the paper's PARAMETERS (which it
# does exactly), and the cohort median must track the closed form. This is the
# claim that can go red on a mis-transcription; the observed-AUC gap above is
# recorded as a deviation and deliberately NOT gated.
med_auc <- median(chk_mb$auclast)
stopifnot(abs(100 * (med_auc - 3 * dose / cl_paper(median(subj$CYSC),
                                                   median(subj$WT))) /
                (3 * dose / cl_paper(median(subj$CYSC), median(subj$WT)))) < 10)
```

# Dosing-regimen simulation (Figure 5 / Table 4)

Because AUC at steady state is exactly (daily dose) / CL for this linear model,
and CL is log-normal with omega = 0.31, the probability of target attainment
has a closed form. The paper's target is fAUC/MIC >= 20 with a 50% protein
binding rate, i.e. total AUCss,0-24h >= 40 x MIC.

```{r pta}
omega_cl <- 0.31
pta <- function(daily_mg, mic, cysc, wt) {
  # P(AUC >= 40*MIC) = P(CL <= daily/(40*MIC)) for log-normal CL
  pnorm(log(daily_mg / (40 * mic * cl_paper(cysc, wt))) / omega_cl)
}

regimens <- tibble::tribble(
  ~regimen,      ~daily_mg,
  "0.5 MU q12h",  44,
  "0.5 MU q8h",   66,
  "0.75 MU q12h", 66,
  "1 MU q12h",    88,
  "0.75 MU q8h",  99,
  "1 MU q8h",     132
)

pta_tab <- tidyr::expand_grid(
  regimens,
  WT   = c(50, 65, 80),
  CYSC = c(1.05, 1.40, 2.31, 3.07, 5.11),
  MIC  = c(0.25, 0.5, 1)
) |>
  dplyr::mutate(PTA = 100 * pta(daily_mg, MIC, CYSC, WT))

ggplot(pta_tab, aes(factor(CYSC), PTA,
                    colour = reorder(regimen, daily_mg),
                    group = regimen)) +
  geom_hline(yintercept = 90, linetype = "dashed", colour = "grey40") +
  geom_line() + geom_point(size = 1) +
  facet_grid(MIC ~ WT, labeller = label_both) +
  labs(
    x = "Cystatin C (mg/L)", y = "PTA, fAUC/MIC >= 20 (%)",
    colour = "Regimen",
    title = "Probability of target attainment by weight, cystatin C and MIC",
    caption = paste(
      "Replicates Figure 5 of Huang 2025. Rows are MIC (mg/L), columns are",
      "body weight (kg). Dashed line is the 90% PTA criterion."
    )
  ) +
  theme(legend.position = "bottom")
```

```{r pta-claims}
# Each row is a claim Huang 2025 makes in prose, re-derived from the closed
# form above. The closed form is deterministic (no random draw), so a failure
# here is a real disagreement with the paper and not cohort noise.
#
# `Deviation = TRUE` marks a claim the packaged model reproducibly does NOT
# reproduce. Those rows stay visible in the table but are excluded from the
# gate, rather than the gate being widened until they pass.
claim <- function(text, achieved, pass, deviation = FALSE) {
  tibble::tibble(Claim = text, Achieved = achieved,
                 Pass = pass, Deviation = deviation)
}

worst_mic1     <- max(pta_tab$PTA[pta_tab$MIC == 1])
worst_80_mic05 <- max(pta_tab$PTA[pta_tab$WT == 80 & pta_tab$MIC == 0.5])

# Cystatin C at which a given daily dose crosses 90% PTA, inverting the
# closed form above. (Re-used by the Table 4 section below.)
thr_cysc <- function(daily_mg, mic, wt) {
  2.31 * (3.69 * (wt / 65)^1.03 /
            (daily_mg / (40 * mic * exp(qnorm(0.9) * omega_cl))))^(1 / 0.626)
}

# Table 4, 50 kg row: the paper recommends 1.5 MU daily (0.5 MU q8h) for
# CysC 3.07-5.11 at MIC 0.25, i.e. its 90%-PTA threshold is the 3.07 grid
# point. Consistency therefore means the closed-form threshold falls in the
# grid step ending at 3.07.
thr_50 <- thr_cysc(66, 0.25, 50)

# Structural claims: PTA rises with dose, rises with CysC, falls with WT.
# These are large effects spanning the whole grid, not near-zero slopes.
mono_dose <- with(
  dplyr::filter(pta_tab, WT == 65, CYSC == 2.31, MIC == 0.25),
  cor(daily_mg, PTA, method = "spearman")
)
mono_wt <- with(
  dplyr::filter(pta_tab, regimen == "1 MU q8h", CYSC == 2.31, MIC == 0.25),
  cor(WT, PTA, method = "spearman")
)
mono_cysc <- with(
  dplyr::filter(pta_tab, regimen == "1 MU q8h", WT == 65, MIC == 0.25),
  cor(CYSC, PTA, method = "spearman")
)

claims <- dplyr::bind_rows(
  claim("PTA increases with daily dose",
        sprintf("Spearman rho = %+.2f", mono_dose), mono_dose > 0.9),
  claim("PTA decreases with body weight",
        sprintf("Spearman rho = %+.2f", mono_wt), mono_wt < -0.9),
  claim("PTA increases with cystatin C",
        sprintf("Spearman rho = %+.2f", mono_cysc), mono_cysc > 0.9),
  claim("At 50 kg, 1.5 MU daily reaches 90% PTA only for CysC >= 3.07 mg/L (MIC 0.25)",
        sprintf("closed-form threshold CysC = %.2f mg/L", thr_50),
        thr_50 >= 2.31 && thr_50 <= 3.07, deviation = TRUE),
  claim("No regimen reaches 90% PTA at MIC = 1 mg/L",
        sprintf("max PTA = %.1f%% (1 MU q8h, 50 kg, CysC 5.11)", worst_mic1),
        worst_mic1 < 90, deviation = TRUE),
  claim("At 80 kg, no regimen reaches 90% PTA at MIC = 0.5 mg/L",
        sprintf("max PTA = %.1f%%", worst_80_mic05),
        worst_80_mic05 < 90, deviation = TRUE)
)

claims |>
  dplyr::mutate(
    Status = dplyr::case_when(
      Pass                   ~ "reproduced",
      Deviation              ~ "KNOWN DEVIATION (not gated)",
      TRUE                   ~ "FAIL"
    )
  ) |>
  dplyr::select(Claim, Achieved, Status) |>
  knitr::kable(caption = paste(
    "Huang 2025 Monte Carlo claims re-derived from the packaged model.",
    "The three structural claims reproduce; the three absolute PTA claims do",
    "not, all in the same direction -- see the note below."
  ))

# Gate only the claims that are not flagged deviations, and assert the
# deviations really are deviations (so this cannot silently become vacuous).
stopifnot(nrow(claims) == 6L)
stopifnot(all(claims$Pass[!claims$Deviation]))
stopifnot(!any(claims$Pass[claims$Deviation]))
```

The three failures share one direction: **the closed form implied by the
paper's own parameters and its own stated target is more optimistic than the
paper's Monte Carlo**, by roughly one cystatin-C grid step. A partial
mechanism is available - if the simulated AUCs were reduced by the same
trapezoidal truncation that affects the observed AUCs (the 0.81 factor
measured above), the maximum PTA at MIC = 1 mg/L falls from
`r sprintf("%.1f", worst_mic1)`% to about 92%, much closer to the paper's
"all regimens fail" - but it does not fully close the gap, so the paper's
simulation evidently carried variability its Methods do not describe. The
*structural* conclusions (higher doses needed at higher weight and lower
cystatin C; failure at raised MIC) reproduce; the exact PTA percentages
should not be quoted from this model as if they were the paper's.

## Table 4 recommendations: a partial reproduction

Table 4 lists the recommended regimen per weight / cystatin-C stratum at
MIC = 0.25 mg/L. The closed form above reproduces the *structure* of that table
- the ordering of regimens, and the direction of every covariate effect - but
its 90%-PTA thresholds sit roughly one cystatin-C grid step **below** the
paper's, i.e. the closed form is more optimistic than the paper's Monte Carlo.

```{r table4}
# thr_cysc() is defined in the claims chunk above.
t4 <- tidyr::expand_grid(regimens, WT = c(50, 65, 80)) |>
  dplyr::mutate(
    cysc_90 = thr_cysc(daily_mg, 0.25, WT),
    note = dplyr::case_when(
      cysc_90 < 1.05 ~ "attained at all CysC in 1.05-5.11",
      cysc_90 > 5.11 ~ "never attained in 1.05-5.11",
      TRUE           ~ sprintf("attained for CysC >= %.2f", cysc_90)
    )
  ) |>
  dplyr::arrange(WT, daily_mg)

t4 |>
  dplyr::select(WT, regimen, daily_mg, note) |>
  dplyr::rename(
    "WT (kg)"           = WT,
    "Regimen"           = regimen,
    "Daily dose (mg)"   = daily_mg,
    "90% PTA at MIC 0.25 mg/L" = note
  ) |>
  knitr::kable(caption = paste(
    "Model-derived counterpart to Table 4 of Huang 2025.",
    "Compare with the paper's stated thresholds -- see the note below."
  ))
```

For the 65 kg stratum the paper states that 0.5 MU q8h / 0.75 MU q12h attain
the target at CysC >= 3.07, 0.75 MU q8h at >= 2.31 and 1 MU q8h at >= 1.4
mg/L; the closed form gives
`r sprintf("%.2f", thr_cysc(66, 0.25, 65))`,
`r sprintf("%.2f", thr_cysc(99, 0.25, 65))` and
`r sprintf("%.2f", thr_cysc(132, 0.25, 65))` mg/L respectively. The offset is
systematic (the paper is more conservative at every dose) and no single
required-AUC threshold reconciles all of the paper's grid points at
omega = 0.31, so the paper's Monte Carlo evidently carried variability beyond
the IIV on CL. This is recorded as a deviation, not gated.

# Assumptions and deviations

* **The WT effect on CL is a POWER term, not an exponential term** - resolving
  a conflict inside the paper. The Results narrative states "CYSC and WT are
  included on CL in power form and exponential form, respectively", but the
  paper's own final-model equation is
  `CL = 3.69 * (CYSC/2.31)^-0.626 * (WT/65)^1.03`, with WT unambiguously
  raised to a power. The equation was confirmed character-for-character
  against the publisher's native rendering of the equation block
  (`10.1177_00368504251325334-eq2.jpg` from the EuropePMC
  `supplementaryFiles` endpoint), so this is a printed value and not a
  digitisation. The exponential reading is also arithmetically impossible:
  `exp(1.03 * WT/65)` would multiply CL by 2.80 at the reference weight rather
  than 1, contradicting the tabulated typical CL of 3.69 L/h, and
  `exp(1.03 * (WT - 65))` would inflate CL about 5e6-fold across the 50-80 kg
  simulation range. Per the standing precedence rule (printed equation
  outranks narrative prose), power form is used. The word "exponential" is
  read as a slip for "exponent".

* **`IIV_CL`, `IIV_V2` and `RSV` in Table 3 are omega standard deviations, not
  variances.** Table 3 gives them as percentages in the final-model column
  (31%, 54.5%, 20.3%) and as bare numbers in the bootstrap-median column
  (0.311, 0.557, 0.193). The bare numbers reproduce the percentages directly,
  which fixes the scale: reading 0.311 as a *variance* would imply a 60.4% CV
  on clearance, double the printed 31%, and would make the bootstrap median
  inconsistent with a final estimate the paper describes as "close". The
  residual row is the clincher - a 0.193 log-scale SD is a 19.3% CV, matching
  the printed 20.3% within bootstrap noise. `ini()` takes variances, so each
  SD is squared in the model file.

* **Fixed allometric scaling was rejected by the authors and is not used.**
  The Methods describe a fixed allometric model `P = theta_p * (WT/70)^K`
  (K = 0.75 on CL/Q, 1 on V1/V2), but the Results state it "was not suitable
  for our data, so it was not considered finally" and Table 2 confirms it
  (OFV 958.59 -> 954.37, p > 0.05, Reserve = NO). The retained WT term is an
  *estimated* exponent of 1.03 on CL only, centred on the cohort median 65 kg
  - not 0.75 on 70 kg. Weight does not enter V1, V2 or Q.

* **Centring constants are cohort medians, not means.** CYSC is centred on
  2.31 mg/L and WT on 65 kg, per the Results sentence "The values 2.31 and 65
  were the medians for CYSC and WT". Table 1's *means* are 2.87 mg/L and
  68.6 kg; substituting them would bias clearance.

* **The published model over-predicts the paper's own observed exposures, and
  this is not tuned away.** At the cohort-mean covariates the printed equation
  gives CL = 3.41 L/h, so 66 mg daily must yield AUCss,0-24h = 19.4 mg*h/L -
  yet the paper reports an observed AUCss,0-24h of 12.51 +/- 6.41 mg*h/L and a
  Css,min of 0.30 +/- 0.22 mg/L against a model trough near 0.6 mg/L. The gap
  is a property of the publication, not of this transcription: the parameter
  block is internally exact (typical CL reproduces 3.69 L/h to machine
  precision, and the steady-state mass-balance identity holds to <1%).
  A quantified part of the discrepancy is the paper's sampling window - AUC
  was computed trapezoidally from samples at 0, 0.5, 1, 2, 4 and 6 h inside an
  **8-h** dosing interval, so the 6-8 h tail was never measured, which alone
  accounts for a factor of about 0.81. The remainder is unexplained by the
  reported model; contributing candidates are the mixed dose levels (10% of
  subjects received 1.0 MU daily and 5% received 2.0 MU), the fact that
  AUCss,0-24h was obtainable for only 16 of 20 subjects, and possible
  non-steady-state troughs. Users comparing this model against observed
  colistin sulfate troughs in CRRT should expect it to run high.

* **Table 4 / Figure 5 reproduce structurally but not numerically.** The
  paper's 90%-PTA cystatin-C thresholds are about one grid step above the
  closed-form values implied by its own model plus its stated target
  (fAUC/MIC >= 20, 50% protein binding, i.e. AUC >= 40 x MIC). No single
  required-AUC threshold reconciles all of the paper's grid points at
  omega = 0.31, so the paper's Monte Carlo evidently included variability
  beyond the IIV on clearance (it is not described in the Methods). The
  qualitative conclusions - failure at MIC >= 0.5-1 mg/L, higher doses needed
  at higher weight and lower cystatin C - reproduce exactly.

* **Covariate distributions are assumed, not published.** Individual data are
  not available. WT is drawn normally and CYSC log-normally with the Table 1
  moments, CYSC truncated to the 1.05-5.11 mg/L range quoted in the Figure 5
  note. No WT/CYSC correlation is imposed; the paper reports none.

* **Only the modal regimen is simulated.** The virtual cohort uses 0.5 MU q8h
  (1.5 MU daily) infused over 0.5 h, which 85% and 90% of subjects
  respectively received. The 2-h infusion (10%), the q12h interval (15%), the
  50% who received a loading dose, and the three subjects who additionally
  received nebulized colistin sulfate are not simulated. Loading dose does not
  affect steady-state AUC for this linear model.

* **CVVHDF flow settings are not covariates.** BFR, DFR, RFR and UFR were all
  screened; UFR reached the forward-inclusion criterion on CL in isolation
  (Table 2, model 2) but was dropped in the second forward step (model 6,
  p > 0.05). The authors attribute the absence of a flow effect to a plateau
  at the high flow rates used. The model therefore carries no dependence on
  CRRT intensity, and should not be extrapolated to substantially different
  CVVHDF settings than the 25-35 mL/kg/h reported.

* **Total bilirubin was significant but is not in the final model.** TBIL
  reached significance in forward selection (Table 2, models 3, 7 and 9) but
  was removed in backward elimination (model 12, p > 0.01). Only CYSC and WT
  are retained, consistent with the paper's final-model equation.

* **All values are from the paper's text and tables** (Table 1, Table 2,
  Table 3, the Results final-model equation, and supplement Table S1). No
  parameter was digitised from a figure, obtained by correspondence, or
  carried from another publication.
