---
title: "Brepocitinib (Hughes 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Brepocitinib (Hughes 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Hughes_2022_brepocitinib"))$reference`
* Description: `r rxode2::rxode(readModelDb("Hughes_2022_brepocitinib"))$description`
* Article: <https://doi.org/10.1002/cpdd.1163>

Brepocitinib (PF-06700841) is an oral tyrosine kinase 2 (TYK2) and Janus kinase
1 (JAK1) inhibitor. Hughes 2022 is the first published population
pharmacokinetic analysis of the programme. A later and larger analysis of the
same molecule, pooling nine studies and 775 subjects into a two-compartment
model, is packaged separately as `Maleki_2024_brepocitinib`.

# Population

The analysis pooled 5541 plasma brepocitinib observations from 379 subjects in
five trials: three phase 1 studies in healthy participants (NCT02310750, the
first-in-human single- and multiple-ascending-dose and food-effect study;
NCT03236493, a multiple-dose study in healthy Japanese participants; and
NCT03656952, a concentration-QT study) and two phase 2 patient studies
(NCT02969018 in moderate-to-severe plaque psoriasis and NCT02974868 in
moderate-to-severe alopecia areata).

Baseline demographics (Hughes 2022 Table 2): median age 42 years (range 18-75),
median body weight 83 kg (range 45-204), median creatinine clearance 124 mL/min
(range 58-279). 69% of subjects were male. Race was 82% White, 8% Asian, 5%
African American and 6% Other. 24% were healthy participants, 55% had plaque
psoriasis and 20% had alopecia areata. 83% received tablets and 17% the oral
suspension used in the first-in-human study.

Doses spanned 1-200 mg. The phase 1 studies contributed single doses of 1, 3,
10, 30, 100 and 200 mg and multiple doses of 10, 30, 100 and 175 mg once daily
or 50 mg twice daily; the phase 2 studies contributed 30 or 60 mg once-daily
induction followed by 10 or 30 mg once-daily maintenance. 10.8% of observations
were missing or below the 0.2 ng/mL limit of quantification and were treated as
missing; an M3 sensitivity analysis moved only the CL/F interindividual
variability.

The same information is available programmatically via
`readModelDb("Hughes_2022_brepocitinib")()$population`.

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Hughes_2022_brepocitinib.R` carries an in-file
comment naming its origin. They are collected here for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` (ka) | 3.46 1/h (95% CI 3.04-3.88) | Table 3, "First-order absorption rate constant"; Equation 7 |
| `lcl` (CL/Frel) | 18.7 L/h (16.9-20.5) | Table 3, "Apparent clearance (CL/F)"; Equation 9 |
| `lvc` (Vc/Frel) | 136 L (124-148) | Table 3, "Apparent volume of distribution (Vc/F)"; Equation 10 |
| `ltlag` (Alag) | 0.240 h (0.234-0.246) | Table 3, "Absorption lag"; Equation 6 |
| `lfdepot` (Frel) | 1 (structural anchor, fixed) | Equation 8, the leading `1 *`; no Table 3 row |
| `e_wt_cl` | 0.750 (fixed) | Table 3, "Effect of weight on CL/F (70 kg reference)"; Equation 9 |
| `e_wt_vc` | 1.00 (fixed) | Table 3, "Effect of weight on Vc/F (70 kg reference)"; Equation 10 |
| `e_form_suspension_tlag` | -1.00 (fixed) | Table 3, "Effect of suspension formulation on Alag"; Equation 6 |
| `e_fed_highfat_ka` | -0.699 (-0.899 to -0.499) | Table 3, "Effect of high-fat meal on ka"; Equation 7 |
| `e_race_asian_cl` | -0.243 (-0.363 to -0.123) | Table 3, "Effect of Asian subjects on CL/F"; Equation 9 |
| `e_fed_highfat_fdepot` | -0.283 (-0.389 to -0.177) | Table 3, "Effect of high-fat meal on Frel"; Equation 8 |
| `e_dose_high_fdepot` | 0.351 (0.194 to 0.508) | Table 3, "Effect of dose on Frel (dose >= 175 mg)"; Equation 8 |
| `var(etalcl)` | 0.6084 = 0.780^2 | Table 3, "omega CL/F (% CV)" = 78.0 (57.5-98.5) |
| `var(etalvc)` | 0.366025 = 0.605^2 | Table 3, "omega Vc/F (%CV)" = 60.5 (27.3-93.5) |
| `cov(etalcl, etalvc)` | 0.358644 = 0.760 x 0.780 x 0.605 | Table 3, "rho CL/F-Vc/F" = 0.760 (0.403-1.12) |
| `expSdHv` | 0.527 (0.444-0.610) | Table 3, "Proportional RUV (phase 1;CV)"; Equation 2 |
| `expSdPat` | 0.875 (0.812-0.938) | Table 3, "Proportional RUV (phase 2;CV)"; Equation 2 |
| Categorical covariate form `P * (1 + theta * COV)` | n/a | Equation 4 |
| Continuous covariate form `P * (COV/COVref)^theta` | n/a | Equation 5 |
| One-compartment ODE with first-order absorption | n/a | Base Model and Discussion; Equations 6-10 |
| Log-transform-both-sides proportional residual error | n/a | Equation 2 |

## Reading the reported "% CV"

The interindividual-variability rows of Table 3 are labelled `omega CL/F (% CV)`
and `omega Vc/F (%CV)`, so the tabulated quantity is omega itself expressed as a
percentage, not the log-normal coefficient of variation
`sqrt(exp(omega^2) - 1)`. Two independent confirmations sit in the same table.
First, the residual-error rows use the identical `CV` label for a quantity
Equation 2 defines unambiguously as a log-scale standard deviation:
`ln(DV) = ln(IPRED) + theta_pro * eps` with `Var(eps)` fixed at 1, and Table 3
reports `theta_pro` = 0.527. Second, the reported 95% confidence intervals are
exactly symmetric about the point estimates on this scale (78.0 with 57.5 to
98.5; 60.5 with 27.3 to 93.5), as an SD-scale interval would be. The model file
therefore uses `omega_CL/F = 0.780` and `omega_Vc/F = 0.605`. For reference, the
equivalent true log-normal CVs are 91.5% and 66.3%.

# Deterministic covariate checks

The paper's quantitative claims are all statements about typical values, so they
are checked here on random-effect-free (`zeroRe`) solves where each ratio is an
exact identity of the published equations rather than a draw from a simulated
cohort.

```{r scenarios}
mod <- readModelDb("Hughes_2022_brepocitinib")
mod_typ <- rxode2::zeroRe(mod)

# Observation grid: 0.01 h out to 6 h so that Tmax (and the 0.240 h absorption
# lag) resolve cleanly, then 0.1 h out to 48 h -- about 9.5 elimination
# half-lives -- so that lambda-z and the AUC extrapolation are well determined.
obs_times <- sort(unique(c(seq(0, 6, by = 0.01), seq(6, 48, by = 0.1))))

# One scenario = one subject. Doses go to the named ODE state `depot` and
# observations to the named ODE state `central`; rxode2 returns the algebraic
# observable Cc alongside, so an algebraic observable must never be named as a
# compartment here -- doing so injects a slot and renumbers the ODE states.
make_scenario <- function(id, scenario, dose, wt = 70, asian = 0, fed = 0,
                          tablet = 1, dose_high = 0, healthy = 1,
                          n_doses = 1L, tau = 24, times = obs_times) {
  dose_times <- (seq_len(n_doses) - 1L) * tau
  ev <- dplyr::bind_rows(
    data.frame(time = dose_times, amt = dose, evid = 1L, cmt = "depot"),
    data.frame(time = max(dose_times) + times, amt = NA_real_,
               evid = 0L, cmt = "central")
  )
  ev$id <- id
  ev$scenario <- scenario
  ev$WT <- wt
  ev$RACE_ASIAN <- asian
  ev$FED_HIGHFAT <- fed
  ev$FORM_TABLET <- tablet
  ev$DOSE_HIGH <- dose_high
  ev$DIS_HEALTHY <- healthy
  ev[order(ev$time, -ev$evid), ]
}

# Reference individual per Hughes 2022 Figure 1: 70 kg, non-Asian, tablet,
# dosed below 175 mg and without regard to food.
ev_typ <- dplyr::bind_rows(
  make_scenario(1L, "30 mg tablet",              dose = 30),
  make_scenario(2L, "30 mg tablet, Asian",       dose = 30, asian = 1),
  make_scenario(3L, "30 mg tablet, high-fat",    dose = 30, fed = 1),
  make_scenario(4L, "30 mg suspension",          dose = 30, tablet = 0),
  make_scenario(5L, "30 mg tablet, 100 kg",      dose = 30, wt = 100),
  make_scenario(6L, "200 mg tablet",             dose = 200, dose_high = 1),
  make_scenario(7L, "200 mg tablet, linear",     dose = 200, dose_high = 0)
)

sim_typ <- rxode2::rxSolve(
  mod_typ, events = ev_typ, keep = "scenario", omega = NA, returnType = "data.frame"
)
str(sim_typ[, c("id", "time", "Cc", "scenario")])
```

```{r nca-typical}
sim_nca <- sim_typ |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, scenario)

# Guarantee a time-zero row per subject; brepocitinib is given orally, so a
# pre-dose concentration of 0 is the correct anchor for AUC0-*.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, scenario) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, scenario, time, .keep_all = TRUE) |>
  dplyr::arrange(id, scenario, time)

dose_df <- ev_typ |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, scenario)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | scenario + id,
                             concu = "ng/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | scenario + id, doseu = "mg")

intervals_sd <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

nca_typ <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals_sd)
)

nca_wide <- as.data.frame(nca_typ) |>
  dplyr::select(scenario, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

get_val <- function(scen, param) {
  nca_wide[[param]][match(scen, nca_wide$scenario)]
}

knitr::kable(
  nca_wide |>
    dplyr::rename(
      "Scenario"          = scenario,
      "Cmax (ng/mL)"      = cmax,
      "Tmax (h)"          = tmax,
      "AUC0-inf (ng*h/mL)" = aucinf.obs,
      "t1/2 (h)"          = half.life
    ),
  digits = 3,
  caption = "Typical-value (zero random effect) NCA for each covariate scenario."
)
```

## The NCA instrument agrees with the closed form

Before the published claims are checked, the NCA itself is checked against the
analytic solution of the model, so that a discrepancy below can be attributed to
the transcription rather than to the measurement.

```{r instrument-check}
cl_ref  <- 18.7                      # L/h, Table 3
vc_ref  <- 136                       # L,   Table 3
ka_ref  <- 3.46                      # 1/h, Table 3
tlag    <- 0.240                     # h,   Table 3
kel_ref <- cl_ref / vc_ref

# For a one-compartment model with complete first-order absorption,
# AUC0-inf = Dose * Frel / CL, exactly, independent of ka and of the lag.
auc_closed_form <- 30 * 1 / cl_ref * 1000       # mg/(L/h) -> ng*h/mL
hl_closed_form  <- log(2) * vc_ref / cl_ref
tmax_closed_tab <- log(ka_ref / kel_ref) / (ka_ref - kel_ref) + tlag

instrument <- tibble::tibble(
  Quantity = c("AUC0-inf, 30 mg tablet (ng*h/mL)", "t1/2 (h)", "Tmax, tablet (h)"),
  `Closed form` = c(auc_closed_form, hl_closed_form, tmax_closed_tab),
  PKNCA = c(get_val("30 mg tablet", "aucinf.obs"),
            get_val("30 mg tablet", "half.life"),
            get_val("30 mg tablet", "tmax"))
) |>
  dplyr::mutate(`% diff` = (PKNCA - `Closed form`) / `Closed form` * 100)

knitr::kable(instrument, digits = 4,
             caption = "PKNCA against the analytic solution of the same model.")

# The solve is deterministic, so these are numerical-integration differences
# only: no cohort is drawn and nothing here varies with the solver thread count.
# Tmax is resolved only to the 0.01 h observation grid, hence its own bound.
stopifnot(
  abs(instrument$`% diff`[instrument$Quantity != "Tmax, tablet (h)"]) < 1,
  abs(get_val("30 mg tablet", "tmax") - tmax_closed_tab) < 0.02
)
```

## Published covariate claims

```{r covariate-claims}
auc_ref     <- get_val("30 mg tablet", "aucinf.obs")
auc_asian   <- get_val("30 mg tablet, Asian", "aucinf.obs")
auc_fed     <- get_val("30 mg tablet, high-fat", "aucinf.obs")
auc_wt100   <- get_val("30 mg tablet, 100 kg", "aucinf.obs")
auc_200_hi  <- get_val("200 mg tablet", "aucinf.obs")
auc_200_lin <- get_val("200 mg tablet, linear", "aucinf.obs")

tmax_tab  <- get_val("30 mg tablet", "tmax")
tmax_susp <- get_val("30 mg suspension", "tmax")
tmax_fed  <- get_val("30 mg tablet, high-fat", "tmax")

# ka is not an NCA parameter, so the high-fat absorption-rate claim is checked
# through the Tmax it implies, computed in closed form from the model's own ka.
ka_fed          <- ka_ref * (1 - 0.699)
tmax_closed_fed <- log(ka_fed / kel_ref) / (ka_fed - kel_ref) + tlag

claims <- tibble::tribble(
  ~Claim, ~Source, ~Published, ~Achieved,
  "CL/F 24.3% lower in Asian participants (AUC ratio)",
  "Abstract; Table 3; Equation 9",
  1 / (1 - 0.243), auc_asian / auc_ref,

  "Relative bioavailability 28.3% lower with a high-fat meal (AUC ratio)",
  "Abstract; Table 3; Equation 8",
  1 - 0.283, auc_fed / auc_ref,

  "Relative bioavailability 35.1% higher at doses >= 175 mg (AUC ratio)",
  "Abstract; Table 3; Equation 8",
  1 + 0.351, auc_200_hi / auc_200_lin,

  "Allometric exponent 0.75 on CL/F: AUC ratio at 100 vs 70 kg",
  "Table 3; Equation 9",
  (100 / 70)^-0.75, auc_wt100 / auc_ref,

  "Absorption lag of 0.240 h for the tablet and none for the suspension (h)",
  "Table 3; Equation 6",
  0.240, tmax_tab - tmax_susp,

  "Absorption rate 69.9% lower with a high-fat meal (implied Tmax, h)",
  "Abstract; Table 3; Equation 7",
  tmax_closed_fed, tmax_fed,

  "Elimination half-life of 5.04 h in the reference population (h)",
  "Results, Final Model",
  5.04, get_val("30 mg tablet", "half.life")
) |>
  dplyr::mutate(
    `% diff` = (Achieved - Published) / Published * 100,
    Pass     = abs(`% diff`) < 1.5
  )

knitr::kable(claims, digits = 4,
             caption = "Published quantitative claims against the packaged model.")

# Every row is a deterministic identity of Equations 6-10 evaluated through a
# numerical solve and a trapezoidal NCA, so the only tolerance needed is for
# integration error. 1.5% leaves headroom over the realised differences while
# still failing on a mis-transcribed coefficient, every one of which moves its
# row by at least several percent (the smallest published effect here, the
# 0.75 allometric exponent, moves AUC by 22% between 70 and 100 kg).
stopifnot(all(claims$Pass))
```

## Comparison against published NCA

Hughes 2022 does not publish a noncompartmental analysis of its own pooled data
set; the model-derived reference half-life of 5.04 h is the one point estimate
the Results give. The Introduction additionally quotes the first-in-human
noncompartmental analysis, whose median time to maximum concentration was "1
hour or less" after the oral suspension.

```{r nca_compare}
published <- tibble::tribble(
  ~scenario,           ~PPTESTCD,   ~PPORRES,
  "30 mg tablet",      "half.life", 5.04,
  "30 mg suspension",  "half.life", 5.04,
  "30 mg suspension",  "tmax",      1.00
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_typ,
  reference     = published,
  by            = "scenario",
  units         = c(half.life = "h", tmax = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated vs. published values. * differs from reference by >20%.",
  align   = c("l", "l", "r", "r", "r")
)
attr(cmp, "footnote")
```

Both half-life rows reproduce the published 5.04 h. The suspension Tmax row
compares against a value the source states as an upper bound ("1 hour or less")
rather than a point estimate, so it is shown for orientation rather than as a
strict target; the model's 0.97 h sits just inside it.

# Replicating Figure 2: steady-state profiles

Hughes 2022 Figure 2 shows steady-state concentration-time profiles simulated
for 2000 individuals at the reference covariate values -- a 70 kg White
participant taking 10, 30 or 60 mg brepocitinib tablets once daily without
regard to food -- summarised as the median and the 5th/95th, 10th/90th and
25th/75th percentile ribbons of the individual predicted concentrations. The
cohort here is 200 subjects per dose arm, the per-arm cap for these vignettes;
the percentile envelope is driven by the published omega matrix, not by the
cohort size.

```{r cohort}
# set.seed() seeds R's RNG, not rxode2's; rxode2 partitions its streams per
# solver thread, so this cohort differs between a 2-thread CI runner and a
# 16-thread workstation. Every assertion below is written to hold for any
# cohort the published omega matrix can produce.
set.seed(20220001)

n_arm  <- 200L
tau    <- 24
n_days <- 10L

ss_times <- seq(0, tau, by = 0.25)

ev_ss <- dplyr::bind_rows(
  lapply(seq_along(c(10, 30, 60)), function(k) {
    d <- c(10, 30, 60)[k]
    dplyr::bind_rows(lapply(seq_len(n_arm), function(i) {
      make_scenario(
        id       = (k - 1L) * n_arm + i,
        scenario = paste0(d, " mg once daily"),
        dose     = d,
        n_doses  = n_days,
        tau      = tau,
        healthy  = 0L,           # phase 2 patients; affects the RUV only
        times    = ss_times
      )
    }))
  })
)

# Duplicate subject IDs across arms would silently collapse into one subject.
stopifnot(
  !anyDuplicated(ev_ss[, c("id", "time", "evid")]),
  dplyr::n_distinct(ev_ss$id) == 3L * n_arm
)

sim_ss <- rxode2::rxSolve(mod, events = ev_ss, keep = "scenario",
                          returnType = "data.frame") |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(time_ss = time - (n_days - 1L) * tau)
```

```{r figure-2, fig.cap = "Replicates Figure 2 of Hughes 2022: steady-state brepocitinib concentration-time profiles at 10, 30 and 60 mg once daily in the reference population."}
sim_ss |>
  dplyr::group_by(scenario, time_ss) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05), Q10 = quantile(Cc, 0.10),
    Q25 = quantile(Cc, 0.25), Q50 = quantile(Cc, 0.50),
    Q75 = quantile(Cc, 0.75), Q90 = quantile(Cc, 0.90),
    Q95 = quantile(Cc, 0.95), .groups = "drop"
  ) |>
  dplyr::mutate(
    scenario = factor(scenario, levels = paste0(c(10, 30, 60), " mg once daily"))
  ) |>
  ggplot(aes(time_ss)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), fill = "steelblue", alpha = 0.20) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), fill = "steelblue", alpha = 0.25) +
  geom_ribbon(aes(ymin = Q25, ymax = Q75), fill = "steelblue", alpha = 0.30) +
  geom_line(aes(y = Q50), colour = "steelblue4", linewidth = 0.8) +
  facet_wrap(~scenario) +
  labs(
    x = "Time after the steady-state dose (h)",
    y = "Brepocitinib concentration (ng/mL)",
    caption = "Replicates Figure 2 of Hughes 2022."
  ) +
  theme_bw()
```

## Steady-state identities

The published equations fix the typical steady-state exposure exactly, so the
deterministic solve is gated tightly and the simulated cohort is only checked
for consistency with it.

```{r steady-state-checks}
ev_ss_typ <- dplyr::bind_rows(
  make_scenario(1L, "10 mg once daily", dose = 10, n_doses = n_days,
                tau = tau, times = ss_times),
  make_scenario(2L, "30 mg once daily", dose = 30, n_doses = n_days,
                tau = tau, times = ss_times),
  make_scenario(3L, "60 mg once daily", dose = 60, n_doses = n_days,
                tau = tau, times = ss_times)
)

sim_ss_typ <- rxode2::rxSolve(mod_typ, events = ev_ss_typ, keep = "scenario",
                              omega = NA, returnType = "data.frame") |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(time_ss = time - (n_days - 1L) * tau)

cavg_typ <- sim_ss_typ |>
  dplyr::group_by(scenario) |>
  dplyr::summarise(
    cavg = sum(diff(time_ss) * (head(Cc, -1) + tail(Cc, -1)) / 2) / tau,
    .groups = "drop"
  )

# Cavg,ss = Dose * Frel / (CL * tau), exactly, for any one-compartment model
# with complete first-order absorption.
cavg_typ$closed_form <- c(10, 30, 60) * 1000 / (cl_ref * tau)
cavg_typ$pct_diff <- (cavg_typ$cavg - cavg_typ$closed_form) /
  cavg_typ$closed_form * 100

knitr::kable(
  cavg_typ |>
    dplyr::rename(
      "Regimen"                 = scenario,
      "Cavg,ss simulated (ng/mL)" = cavg,
      "Cavg,ss closed form (ng/mL)" = closed_form,
      "% diff"                  = pct_diff
    ),
  digits = 3,
  caption = "Typical-value steady-state average concentration against its closed form."
)

# Accumulation ratio for a one-compartment model dosed every tau.
rac_closed <- 1 / (1 - exp(-kel_ref * tau))

cmax_first <- sim_ss_typ |>
  dplyr::filter(scenario == "30 mg once daily") |>
  dplyr::summarise(m = max(Cc)) |>
  dplyr::pull(m)

ev_first <- make_scenario(1L, "30 mg once daily", dose = 30, n_doses = 1L,
                          times = ss_times)
cmax_single <- rxode2::rxSolve(mod_typ, events = ev_first, omega = NA,
                               returnType = "data.frame") |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::summarise(m = max(Cc)) |>
  dplyr::pull(m)

ss_claims <- tibble::tribble(
  ~Claim, ~Published, ~Achieved,
  "Dose proportionality below 175 mg: Cavg,ss(60 mg) / Cavg,ss(30 mg)",
  2, cavg_typ$cavg[3] / cavg_typ$cavg[2],
  "Dose proportionality below 175 mg: Cavg,ss(30 mg) / Cavg,ss(10 mg)",
  3, cavg_typ$cavg[2] / cavg_typ$cavg[1],
  "Accumulation ratio at 24 h dosing implied by a 5.04 h half-life",
  rac_closed, cmax_first / cmax_single
) |>
  dplyr::mutate(`% diff` = (Achieved - Published) / Published * 100)

knitr::kable(ss_claims, digits = 4,
             caption = "Steady-state identities of the packaged model.")

stopifnot(
  abs(cavg_typ$pct_diff) < 1,
  abs(ss_claims$`% diff`) < 1.5
)
```

```{r cohort-consistency}
# The etas are log-normal with median 1, so the MEDIAN individual Cavg,ss must
# sit on the typical value. With omega_CL/F = 0.780 and 200 subjects per arm the
# standard error of that sample median is about 7% on the log scale
# (1.2533 * omega / sqrt(n)), and the three arms here realise +3.8%, +15.7% and
# -3.1% -- ordinary sampling scatter of up to a couple of standard errors.
#
# The gate is therefore set at 40%, not at the realised maximum. rxSetSeed()
# fixes rxode2's stream per solver thread but NOT across thread counts, so CI
# draws a different cohort than any one workstation does; a bound placed just
# above what happened to be observed here is the single most common way these
# vignettes pass locally and fail in CI. 40% is about 5.8 standard errors, so
# no cohort the published omega can produce will trip it, while a
# mis-transcribed clearance, dose or unit -- each of which moves the centre by
# tens of percent, a factor-of-2 slip by 100% -- still will.
cohort_cavg <- sim_ss |>
  dplyr::group_by(scenario, id) |>
  dplyr::summarise(
    cavg = sum(diff(time_ss) * (head(Cc, -1) + tail(Cc, -1)) / 2) / tau,
    .groups = "drop"
  ) |>
  dplyr::group_by(scenario) |>
  dplyr::summarise(median_cavg = median(cavg), .groups = "drop") |>
  dplyr::arrange(match(scenario, paste0(c(10, 30, 60), " mg once daily")))

cohort_cavg$typical <- c(10, 30, 60) * 1000 / (cl_ref * tau)
cohort_cavg$pct_diff <- (cohort_cavg$median_cavg - cohort_cavg$typical) /
  cohort_cavg$typical * 100

knitr::kable(
  cohort_cavg |>
    dplyr::rename(
      "Regimen"                       = scenario,
      "Median individual Cavg,ss (ng/mL)" = median_cavg,
      "Typical-value Cavg,ss (ng/mL)"     = typical,
      "% diff"                        = pct_diff
    ),
  digits = 2,
  caption = "Cohort median against the typical value; log-normal IIV puts them on top of each other in expectation."
)

stopifnot(abs(cohort_cavg$pct_diff) < 40)
```

# Food effect against the published noncompartmental analysis

The Introduction reports that a high-fat meal reduced the first-in-human
noncompartmental AUC and Cmax by approximately 17.7% and 35.7%. The final model
attributes the food effect to a 28.3% reduction in relative bioavailability and
a 69.9% reduction in the absorption-rate constant. The Discussion addresses the
AUC discrepancy directly: the model-based reduction "appeared slightly greater
than that determined by noncompartmental analysis", but the covariate's 95%
confidence interval of 16.9%-39.1% contains the noncompartmental value, so the
two are not in conflict.

```{r food-effect}
cmax_ref <- get_val("30 mg tablet", "cmax")
cmax_fed <- get_val("30 mg tablet, high-fat", "cmax")

food <- tibble::tibble(
  Metric = c("AUC0-inf reduction (%)", "Cmax reduction (%)"),
  `Published NCA (Introduction)` = c(17.7, 35.7),
  `Final model (this vignette)` = c(
    (1 - auc_fed / auc_ref) * 100,
    (1 - cmax_fed / cmax_ref) * 100
  )
)

knitr::kable(food, digits = 1,
             caption = "Food effect: the final model against the cited noncompartmental analysis.")

# Reported as a documented deviation, not gated. The two numbers come from
# different analyses of overlapping but not identical data, and the paper's own
# Discussion explains the AUC gap. Only the model-internal identity -- that the
# AUC reduction equals the fitted 28.3% -- is gated, and it is gated above.
```

# Assumptions and deviations

* **Reading of "% CV".** Table 3 reports the interindividual variabilities as
  `omega ... (% CV)` and the residual errors as `... (CV)`. Both are taken as
  log-scale standard deviations rather than log-normal coefficients of
  variation, for the reasons set out under "Reading the reported % CV" above.
  Had the alternative reading been used, `omega_CL/F` would be 0.690 instead of
  0.780 and `omega_Vc/F` 0.559 instead of 0.605. No result gated in this
  vignette depends on the choice, because every gate is on a typical value.
* **Units of `ka`.** Table 3 and the Abstract print the absorption rate constant
  as "3.46 h". A first-order rate constant has units of 1/h, and 3.46 1/h is
  what reproduces both the reported 5.04 h half-life population and the
  sub-1-hour noncompartmental Tmax; the printed unit is a typographical error.
* **Orientation of the formulation covariate.** Hughes 2022 Equation 6 is
  written on a suspension indicator, `Alag = 0.24 * (1 - 1 * Suspension)`. The
  canonical register column is the tablet-oriented `FORM_TABLET`, so the model
  file derives `Suspension = 1 - FORM_TABLET` inside `model()` and reproduces
  the printed equation unchanged. The companion `Maleki_2024_brepocitinib`
  model uses the same column and orientation.
* **`DOSE_HIGH` threshold.** This model sets `DOSE_HIGH = 1` at 175 mg and
  above, per Equation 8; the companion `Maleki_2024_brepocitinib` model sets it
  above 100 mg/day. Neither analysis studied a dose between 100 and 175 mg, so
  the two encodings select the same subjects. Hughes 2022 is explicit that the
  step function is an artefact of that gap and that the true relationship
  probably rises gradually between the two doses.
* **`DIS_HEALTHY` drives the residual error only.** The paper estimated separate
  proportional residual errors for its phase 1 and phase 2 studies. Every phase 1
  subject was a healthy participant and every phase 2 subject was a patient, and
  the Abstract restates the two values as applying to "healthy volunteers" and
  "patients", which is what licenses the mapping onto this column. Patient status
  was separately screened as a structural covariate on CL/F and on Frel and
  retained in neither.
* **Covariates screened but not retained.** Age, sex, and the patient-type
  effect on CL/F and Frel were evaluated (Table 1) and dropped. They are
  recorded in the model file's `covariatesDataExcluded` metadata rather than
  `covariateData`, since they are not referenced in `model()`. Creatinine
  clearance appears in the Table 2 demographics but was not among the covariates
  evaluated.
* **Race effects other than Asian.** Exploratory analysis found a statistically
  significant effect of "other races" (Native American and Pacific Islander) on
  CL/F, but it was excluded from both the full and final models because at most
  three such subjects had full PK profiles. The packaged model follows the final
  model and omits it.
* **The `eps_res` row of Table 3.** Table 3 lists `eps_res` as 1.00 (fixed).
  This is the NONMEM SIGMA that Equation 2 fixes to 1 so that `theta_pro` is
  identified as the log-scale standard deviation. It has no nlmixr2 counterpart
  and is not carried into `ini()`.
* **No supplement was available.** The Supplemental Information is cited for the
  sampling schedules, the model-development step table, the visual predictive
  checks, the below-limit-of-quantification sensitivity analysis and the final
  NONMEM control stream. It is not included in the open-access deposit. Every
  parameter of the final model is printed in Table 3 and Equations 6-10 of the
  main text, so nothing needed for the model file is missing; the unavailable
  material would only have added diagnostics.
* **Cohort size.** Figure 2 of the paper is based on 2000 simulated
  individuals; the replication here uses 200 per dose arm, the per-arm cap for
  these vignettes. The percentile envelope is set by the published omega matrix
  and is essentially unchanged at either size.
