---
title: "Vancomycin in class 3 obesity, parametric vs nonparametric (Hughes 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vancomycin in class 3 obesity, parametric vs nonparametric (Hughes 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Hughes 2024 is a quality-improvement effort that fitted **two** population PK models to
the *same* vancomycin therapeutic-drug-monitoring dataset from a single institution, in
order to ask whether a parametric or a nonparametric estimation method produces the
better model for model-informed precision dosing (MIPD). Both models are packaged here.

```{r models}
mod_par <- readModelDb("Hughes_2024_vancomycin_parametric")
mod_np  <- readModelDb("Hughes_2024_vancomycin_nonparametric")
```

* Parametric (nlmixr2 SAEM): `r rxode2::rxode(mod_par)$description`
* Nonparametric (Pmetrics NPAG): `r rxode2::rxode(mod_np)$description`
* Citation: `r rxode2::rxode(mod_par)$reference`
* Article: <https://doi.org/10.1097/FTD.0000000000001214>
* Supplement (Supplemental Digital Content 1): <http://links.lww.com/TDM/A753>

The paper also fitted the parametric model a third time in NONMEM as a software
cross-check, and reports that parameter set in Table 2 as well. It is not packaged
again here: the Results state "we present only the results for nlmixr2", Table 3
("Comparison of Final Models") enumerates exactly two final models, and the NONMEM
parameter set is already available in this library as
`modellib("Tong_2026_vancomycin_hughes")`, transcribed from the control stream in the
Tong 2026 supplement. See the Errata below, where the three parametric parameter sets
are compared side by side.

# Population

Both models were fitted to the same 83 patients contributing 272 vancomycin TDM levels
(median 2 per patient, range 1-29; 31 peaks, 115 troughs, 126 random), collected at the
University of Vermont Medical Center between 1 November 2021 and 14 February 2023 and
exported from the InsightRX Nova precision-dosing software. Inclusion required a body
mass index of at least 40 kg/m^2 at some point during treatment and at least one
vancomycin level. Median age was 56.3 years (range 24.2-89.3), median weight 134 kg
(range 91.6-218), median BMI 46.3 kg/m^2 (range 40-70.3), median serum creatinine
0.84 mg/dL (range 0.40-2.51), and the cohort was 42/83 female. Race and ethnicity were
not reported, and the Discussion records as a limitation that patient status, level of
care and indication for vancomycin were not collected. Three levels were removed during
data cleaning; all remaining levels were above the lower limit of quantification.
(Table 1, development-dataset column.)

A separate external validation dataset of 576 patients across 74 US organizations
(1191 levels) was used to compare the parametric model's predictive performance against
four published obesity models, but was not used to estimate any parameter.

The same information is available programmatically via each model's `population`
metadata (`rxode2::rxode(readModelDb("Hughes_2024_vancomycin_parametric"))$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()` entry in
`inst/modeldb/specificDrugs/Hughes_2024_vancomycin_parametric.R` and
`..._nonparametric.R`. The table below collects them in one place. "Table 2 (par.)" and
"Table 2 (np.)" refer to the nlmixr2 (SAEM) and Pmetrics columns of Table 2
respectively; "SDC1 S<n>" refers to section `n` of Supplemental Digital Content 1.

| Equation / parameter | Parametric | Nonparametric | Source location |
|---|---|---|---|
| `lcl` (CL at CrCl 100 mL/min) | `log(5.09)` L/h | `log(6.74)` L/h | Table 2, CL row |
| `lvc` (Vc at FFM 70 kg) | `log(64.9)` L | `log(104)` L | Table 2, Vc row |
| `lq` (Q) | `log(6.36)` L/h | `log(7.5)` L/h | Table 2, Q row |
| `lvp` (Vp at FFM 70 kg) | `log(66.4)` L | `log(142)` L | Table 2, Vp row |
| `e_crcl_cl` (exponent on CrCl/100) | `0.887` | `1.05` | Table 2, u3 row |
| `e_ffm_vc_vp` (exponent on FFM/70) | `fixed(1)` | `fixed(1)` | Table 2, u2 row (FIX) |
| `etalcl` | `0.249^2` | `log(0.222^2+1)` | Table 2, CL BSV (24.9 / 22.2 %CV) |
| `etalvc` | `0.178^2` | `log(0.252^2+1)` | Table 2, Vc BSV (17.8 / 25.2 %CV) |
| `etalq` | absent | `log(0.54^2+1)` | Table 2, Q BSV ("-" / 54 %CV); Table 3 |
| `etalvp` | `0.803^2` | `log(0.350^2+1)` | Table 2, Vp BSV (80.3 / 35.0 %CV) |
| `propSd` | `0.168` | `fixed(0.1)` | Table 2, RUV row |
| `addSd` | `fixed(0.001)` | `fixed(0.5)` | SDC1 S3 / S4 (par.); Table 2 RUVadd (np.) |
| Error combination | quadrature | `combined1()` | SDC1 S4 `W = SQRT(...)`; S5 `#Err` polynomial |
| Covariate model `CLi`, `Vi`, `Qi`, `Vp,i` | n/a | n/a | Table 2 footnote; SDC1 S4 `$PK`; S5 `#Sec` |
| BMI, FFM (Janmahasatian), Cockcroft-Gault on FFM | n/a | n/a | Methods/Data Collection; SDC1 S4 `$PK` |
| Structural model: 2-compartment IV | n/a | n/a | Methods/Model Development; Table 3 |

The allometric exponent `u1` that appears in the Table 2 footnote
(`CLi = CL*(FFM/70)^u1*(CrCL/100)^u3`, `Qi = Q*(FFM/70)^u1`) is **not** part of either
final model: it has no row in Table 2, the Results state for the parametric fit that it
"considerably worsened the model fit when fixed and was estimated to be close to zero
when estimated; therefore, this exponent was excluded" and for the nonparametric fit
that "once CrCl was included as a predictor of CL, the inclusion of the allometric
exponent on CL and Q did not improve the model and was, therefore, removed", Table 3
lists the final covariates as CL~CrCl / Vc~FFM / Vp~FFM only, and the SDC1 S4 `$PK` and
S5 `#Sec` blocks both write Q with no covariate term. Both packaged models therefore
set `u1 = 0`.

# Structural comparison (reproduces Table 3)

Table 3's "Model structure" rows are recovered directly from the two packaged model
objects, which is a check that the encoding carries the paper's one structural
difference between the approaches: the nonparametric fit puts between-subject
variability on all four disposition parameters, the parametric fit on only three.

```{r table3-structure}
struct_row <- function(mod, label) {
  ui <- rxode2::rxode(mod)
  etas <- ui$iniDf$name[!is.na(ui$iniDf$neta1)]
  tibble(
    Approach   = label,
    Structural = "2-compartment",
    Covariates = "CL ~ CrCl; Vc ~ FFM; Vp ~ FFM",
    BSV        = paste(sub("^etal", "", etas), collapse = ", "),
    `Residual error` = ui$predDf$addProp
  )
}

structure_tbl <- bind_rows(
  struct_row(mod_par, "Parametric (nlmixr2)"),
  struct_row(mod_np,  "Nonparametric (Pmetrics)")
)

knitr::kable(structure_tbl, caption = "Reproduces the 'Model structure' rows of Table 3 of Hughes 2024.")

# Table 3 states BSV on "CL, Vc, Vp" (parametric) and "CL, Vc, Q, Vp" (nonparametric).
stopifnot(
  structure_tbl$BSV[structure_tbl$Approach == "Parametric (nlmixr2)"] == "cl, vc, vp",
  structure_tbl$BSV[structure_tbl$Approach == "Nonparametric (Pmetrics)"] == "cl, vc, q, vp"
)
```

# Virtual cohort

Original observed data are not publicly available. The cohort below is drawn to match
the Table 1 development-dataset demographics. Height and BMI are sampled and weight is
derived as `WT = BMI * (HT/100)^2`, which is the same relation the model uses
internally; this reproduces the published median weight without having to sample weight
and height as if they were independent.

```{r cohort}
# set.seed() seeds R's RNG (used for the covariate draws below). It does NOT seed
# rxode2's simulation RNG, whose streams are partitioned per solver thread -- so the
# IIV draws in the VPC section differ between a 2-core CI runner and a 16-thread
# workstation. Every assertion below is written to hold for any cohort the model can
# produce; see pattern 12 of the skill's known-vignette-failure-patterns reference.
set.seed(20240905)
n_subj <- 200L  # per model arm; the skill caps cohorts at 200/arm

clamp <- function(x, lo, hi) pmin(pmax(x, lo), hi)

cohort <- tibble(
  id    = seq_len(n_subj),
  HT    = clamp(rnorm(n_subj, 170, 11),           122,   190),   # Table 1: 170 (122-190) cm
  BMI   = clamp(40 + rlnorm(n_subj, log(6.3), .65), 40,   70.3), # Table 1: 46.3 (40-70.3)
  AGE   = clamp(rnorm(n_subj, 56.3, 15),          24.2,  89.3),  # Table 1: 56.3 (24.2-89.3)
  SEXF  = rbinom(n_subj, 1, 42 / 83),                            # Table 1: 42/83 female
  CREAT = clamp(rlnorm(n_subj, log(0.84), 0.35),  0.40,  2.51)   # Table 1: 0.84 (0.40-2.51)
) |>
  mutate(WT = BMI * (HT / 100)^2)
```

## Covariate machinery check (reproduces Table 1)

The Janmahasatian fat-free-mass equations and the Cockcroft-Gault-on-FFM creatinine
clearance are computed *inside* the model, so they are not visible in `ini()` and are
not covered by the parameter source-trace. Table 1 nevertheless reports the resulting
cohort medians for **both** creatinine-clearance flavours -- 84.1 mL/min on fat-free
mass and 174 mL/min on total body weight -- which is enough to test the derivation
independently. Recomputing them here with the same equations the model body uses is a
direct check on that machinery.

```{r table1-check}
ffm_of  <- function(WT, BMI, SEXF) {
  ifelse(SEXF == 1, 9270 * WT / (8780 + 244 * BMI), 9270 * WT / (6680 + 216 * BMI))
}
crcl_of <- function(size, AGE, SEXF, CREAT) (140 - AGE) * size * 0.85^SEXF / (72 * CREAT)

cohort <- cohort |>
  mutate(
    FFM      = ffm_of(WT, BMI, SEXF),
    CRCL_FFM = crcl_of(FFM, AGE, SEXF, CREAT),
    CRCL_TBW = crcl_of(WT,  AGE, SEXF, CREAT)
  )

table1 <- tibble::tribble(
  ~Characteristic,                       ~Published, ~Simulated,
  "Weight (kg)",                              134,   median(cohort$WT),
  "Height (cm)",                              170,   median(cohort$HT),
  "BMI (kg/m^2)",                            46.3,   median(cohort$BMI),
  "Age (years)",                             56.3,   median(cohort$AGE),
  "Serum creatinine (mg/dL)",                0.84,   median(cohort$CREAT),
  "Creatinine clearance on TBW (mL/min)",     174,   median(cohort$CRCL_TBW),
  "Creatinine clearance on FFM (mL/min)",    84.1,   median(cohort$CRCL_FFM)
) |>
  mutate(`% diff` = 100 * (Simulated - Published) / Published)

knitr::kable(table1, digits = 1,
             caption = "Virtual-cohort medians against the Table 1 development-dataset medians of Hughes 2024.")

# The two creatinine-clearance rows are the load-bearing ones: they exercise the
# Janmahasatian FFM equations and the Cockcroft-Gault substitution. A mis-transcribed
# FFM coefficient or a Cockcroft-Gault written on total body weight moves them by a
# factor of ~2 (the published TBW/FFM ratio is 174/84.1 = 2.07), so a 15% band still
# goes red on any such error while absorbing the sampling noise of a 200-subject draw.
stopifnot(all(abs(table1$`% diff`) < 15))
```

# Typical-value parameters (reproduces Table 2 and the Figure 2 claim)

Evaluating each model at a fixed covariate profile recovers the Table 2 estimates when
that profile sits at the models' reference points (FFM = 70 kg, CrCl = 100 mL/min), and
lets the two approaches be compared at the cohort's median profile. The paper's
Figure 2 / Discussion claim is that "on an average, Pmetrics estimated higher values of
CL and V than nlmixr2".

```{r typical-values}
tv_of <- function(mod, crcl, ffm) {
  th <- rxode2::rxode(mod)$theta
  c(CL = unname(exp(th[["lcl"]]) * (crcl / 100)^th[["e_crcl_cl"]]),
    Vc = unname(exp(th[["lvc"]]) * (ffm / 70)^th[["e_ffm_vc_vp"]]),
    Q  = unname(exp(th[["lq"]])),
    Vp = unname(exp(th[["lvp"]]) * (ffm / 70)^th[["e_ffm_vc_vp"]]))
}

# At the reference profile the covariate terms are 1, so this must return Table 2.
ref_tbl <- bind_rows(
  c(Approach = "Parametric (nlmixr2)",     tv_of(mod_par, crcl = 100, ffm = 70)),
  c(Approach = "Nonparametric (Pmetrics)", tv_of(mod_np,  crcl = 100, ffm = 70))
) |>
  mutate(across(-Approach, as.numeric))

knitr::kable(ref_tbl, digits = 2,
             caption = "Model parameters at the reference profile (FFM = 70 kg, CrCl = 100 mL/min); these are the Table 2 estimates of Hughes 2024.")

# Deterministic: a pure function of the ini() values, so an exact check is correct here.
stopifnot(
  isTRUE(all.equal(as.numeric(ref_tbl[1, c("CL", "Vc", "Q", "Vp")]), c(5.09, 64.9, 6.36, 66.4))),
  isTRUE(all.equal(as.numeric(ref_tbl[2, c("CL", "Vc", "Q", "Vp")]), c(6.74, 104,  7.50, 142 )))
)
```

```{r figure2-claim}
med_ffm  <- median(cohort$FFM)
med_crcl <- median(cohort$CRCL_FFM)

med_tbl <- bind_rows(
  c(Approach = "Parametric (nlmixr2)",     tv_of(mod_par, med_crcl, med_ffm)),
  c(Approach = "Nonparametric (Pmetrics)", tv_of(mod_np,  med_crcl, med_ffm))
) |>
  mutate(across(-Approach, as.numeric))

knitr::kable(med_tbl, digits = 2,
             caption = sprintf(
               "Model parameters at the cohort median profile (FFM = %.1f kg, CrCl = %.1f mL/min).",
               med_ffm, med_crcl))

# Figure 2 / Discussion: "On an average, Pmetrics estimated higher values of CL and V
# than nlmixr2." This is a comparison of two DETERMINISTIC functions of the published
# ini() values at a common covariate profile -- not a race between two noisy simulated
# statistics -- so a strict ordering assertion is legitimate here. It holds at every
# covariate profile, because the nonparametric CL and Vc exceed the parametric ones at
# the reference point and the covariate terms are shared.
stopifnot(
  med_tbl$CL[2] > med_tbl$CL[1],
  med_tbl$Vc[2] > med_tbl$Vc[1]
)
```

# Closed-form validation: AUC0-inf equals Dose / CL

Neither model's source paper reports an NCA table, so there is no published Cmax / AUC /
half-life to compare against. What *is* available is the exact identity that a linear
two-compartment model must satisfy: after a single intravenous dose, `AUC0-inf` equals
`Dose / CL`, where `CL` is computed independently from the published `ini()` values and
the covariate equations. Simulating with the random effects zeroed makes this a
deterministic check of the whole chain -- covariate derivation, parameter scaling, ODE
system, and observation equation -- against a quantity that never touches the solver.

```{r closed-form-events}
# A single median-profile subject per model, dosed 2000 mg as a 2-hour infusion.
ref_subj <- list(WT = 134, HT = 170, AGE = 56.3, SEXF = 0, CREAT = 0.84)
dose_amt <- 2000

# Observations are written on the ODE state "central" -- NEVER on the observable "Cc",
# which would auto-inject a cmt() slot after the ODE states and renumber them.
ev_single <- rxode2::et(amt = dose_amt, dur = 2, cmt = "central") |>
  rxode2::et(seq(0, 480, by = 0.25), cmt = "central")

solve_typical <- function(mod) {
  rxode2::rxSolve(rxode2::zeroRe(mod), ev_single,
                  params = unlist(ref_subj), returnType = "data.frame")
}

sim_single <- bind_rows(
  solve_typical(mod_par) |> mutate(Approach = "Parametric (nlmixr2)",     id = 1L),
  solve_typical(mod_np)  |> mutate(Approach = "Nonparametric (Pmetrics)", id = 2L)
)
```

```{r closed-form-pknca}
sim_nca <- sim_single |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, Approach)

# Guarantee a time = 0 row per (id, Approach). For an IV infusion the pre-dose
# concentration is 0, and PKNCA needs the row to anchor AUC from the interval start.
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, Approach) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, Approach, time, .keep_all = TRUE) |>
  arrange(id, Approach, time)

dose_df <- tibble(id = c(1L, 2L), time = 0,  amt = dose_amt,
                  Approach = c("Parametric (nlmixr2)", "Nonparametric (Pmetrics)"))

conc_obj <- PKNCA::PKNCAconc(as.data.frame(sim_nca), Cc ~ time | Approach + id,
                             concu = "mg/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | Approach + id,
                             doseu = "mg")

intervals <- data.frame(start = 0, end = Inf,
                        cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE)

nca_single <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r closed-form-compare}
# Reference side: the closed-form values, computed from the published ini() numbers and
# the paper's covariate equations WITHOUT touching the solver output, so the comparison
# can actually go red.
bmi_ref  <- ref_subj$WT / (ref_subj$HT / 100)^2
ffm_ref  <- ffm_of(ref_subj$WT, bmi_ref, ref_subj$SEXF)
crcl_ref <- crcl_of(ffm_ref, ref_subj$AGE, ref_subj$SEXF, ref_subj$CREAT)

closed_form <- lapply(
  list("Parametric (nlmixr2)" = mod_par, "Nonparametric (Pmetrics)" = mod_np),
  function(m) {
    p   <- tv_of(m, crcl_ref, ffm_ref)
    kel <- p[["CL"]] / p[["Vc"]]; k12 <- p[["Q"]] / p[["Vc"]]; k21 <- p[["Q"]] / p[["Vp"]]
    b   <- kel + k12 + k21
    lz  <- (b - sqrt(b^2 - 4 * kel * k21)) / 2   # terminal (beta) rate constant
    c(aucinf.obs = dose_amt / p[["CL"]], half.life = log(2) / lz)
  })

reference <- tibble(
  Approach   = names(closed_form),
  aucinf.obs = vapply(closed_form, `[[`, numeric(1), "aucinf.obs"),
  half.life  = vapply(closed_form, `[[`, numeric(1), "half.life")
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_single,
  reference     = reference,
  by            = "Approach",
  params        = c("aucinf.obs", "half.life"),
  units         = c(aucinf.obs = "mg*h/L", half.life = "h"),
  tolerance_pct = 5
)

knitr::kable(cmp, caption = paste(
  "Simulated NCA against the closed-form identities AUC0-inf = Dose/CL and",
  "t1/2 = ln(2)/beta. The reference column is NOT a published NCA table --",
  "Hughes 2024 reports none -- but is derived from the published ini() values.",
  "* marks a difference above 5%."), align = c("l", "l", "r", "r", "r"))
```

```{r closed-form-assert}
# ncaComparisonTable() returns `% diff` as TEXT carrying a "%" suffix and a trailing
# "*" on out-of-tolerance rows, so both must be stripped before a numeric comparison --
# as.numeric("-0.3%") is NA, and all(NA) is NA, which would make this gate un-red-able.
pct <- as.numeric(sub("%$", "", sub("\\*$", "", cmp$`% diff`)))
stopifnot(nrow(cmp) == 4L, !anyNA(pct))   # the gate must have rows to test (pattern 10)

# Deterministic quantities: no IIV, no residual error, one fixed covariate profile.
# The only error sources are trapezoidal integration on a 0.25 h grid and PKNCA's
# lambda-z tail extrapolation -- realised 0.0% on AUC and 0.3% on half-life -- so a 2%
# bound is tight enough to catch a mis-scaled volume, a dropped covariate term or a
# wrong error-model combination, all of which move AUC by tens of percent.
stopifnot(all(abs(pct) < 2))
```

# Simulated dosing under the study protocol

The institutional protocol (Methods, Data Collection) gave patients with BMI >= 40
kg/m^2 a 20 mg/kg loading dose capped at 4000 mg, followed by an intermittent-infusion
maintenance regimen chosen to reach an AUC of 400-600 mg*h/L. The cohort below is dosed
that way -- loading dose, then 1000 mg every 12 h for 7 days -- and simulated through
each model with between-subject variability, giving the visual-predictive-check-style
picture that Figure 1 of the paper shows for each approach.

```{r protocol-events}
tau      <- 12    # h, maintenance dosing interval
n_days   <- 7
rate_max <- 1000  # mg/h; standard maximum vancomycin infusion rate (assumption, see Errata)

make_events <- function(cohort, id_offset = 0L) {
  subj <- cohort |> mutate(id = id + id_offset)

  load_dose <- subj |>
    mutate(time = 0, amt = pmin(20 * WT, 4000), evid = 1, cmt = "central",
           dur = amt / rate_max)

  maint <- subj |>
    tidyr::crossing(time = seq(tau, n_days * 24 - tau, by = tau)) |>
    mutate(amt = 1000, evid = 1, cmt = "central", dur = amt / rate_max)

  obs <- subj |>
    tidyr::crossing(time = seq(0, n_days * 24, by = 0.5)) |>
    mutate(amt = NA_real_, evid = 0, cmt = "central", dur = NA_real_)

  bind_rows(load_dose, maint, obs) |> arrange(id, time, desc(evid))
}

events <- make_events(cohort)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

```{r protocol-simulate}
sim_one <- function(mod, label) {
  # Re-seed immediately before each solve so both arms draw the SAME etas (common
  # random numbers); otherwise the two models would be compared across different
  # virtual patients and the difference between them would be confounded by sampling.
  rxode2::rxSetSeed(42)
  rxode2::rxSolve(mod, events = events,
                  keep = c("WT", "FFM", "CRCL_FFM", "SEXF")) |>
    as.data.frame() |>
    mutate(Approach = label)
}

sim <- bind_rows(
  sim_one(mod_par, "Parametric (nlmixr2)"),
  sim_one(mod_np,  "Nonparametric (Pmetrics)")
)

stopifnot(all(is.finite(sim$Cc)), all(sim$Cc >= 0))
```

```{r figure-1, fig.cap = "Replicates the layout of Figure 1 of Hughes 2024: simulated concentration-time percentiles for each modeling approach."}
sim |>
  group_by(Approach, time) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
            .groups = "drop") |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue4") +
  facet_wrap(~Approach) +
  labs(x = "Time (h)", y = "Vancomycin concentration (mg/L)",
       title = "Simulated VPC by modeling approach",
       subtitle = "20 mg/kg load (max 4000 mg), then 1000 mg q12h; median and 5th-95th percentiles",
       caption = "Replicates the layout of Figure 1 of Hughes 2024.")
```

## Steady-state exposure against the protocol AUC target

```{r steady-state-nca}
ss_start <- (n_days * 24) - tau   # last complete dosing interval

sim_ss <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, Approach)

conc_ss <- PKNCA::PKNCAconc(as.data.frame(sim_ss), Cc ~ time | Approach + id,
                            concu = "mg/L", timeu = "h")

dose_ss <- PKNCA::PKNCAdose(
  as.data.frame(events |> dplyr::filter(evid == 1) |> dplyr::select(id, time, amt)),
  amt ~ time | id, doseu = "mg")

int_ss <- data.frame(start = ss_start, end = ss_start + tau,
                     auclast = TRUE, cmax = TRUE, cmin = TRUE)

nca_ss <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_ss, dose_ss, intervals = int_ss))

ss_tbl <- as.data.frame(nca_ss$result) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  # AUC over one 12 h interval at steady state -> daily AUC is twice that.
  mutate(auc24 = PPORRES * 24 / tau) |>
  group_by(Approach) |>
  summarise(`Median AUC0-24,ss (mg*h/L)` = median(auc24),
            `5th percentile` = quantile(auc24, 0.05),
            `95th percentile` = quantile(auc24, 0.95),
            `% in 400-600 target` = 100 * mean(auc24 >= 400 & auc24 <= 600),
            .groups = "drop")

knitr::kable(ss_tbl, digits = 1, caption = paste(
  "Steady-state daily exposure on the protocol regimen, by modeling approach.",
  "The 400-600 mg*h/L window is the AUC target stated in Methods/Data Collection."))
```

```{r steady-state-assert}
# The paper's headline conclusion is that the two approaches agree clinically:
# "the clinical significance of these distributional differences seemed minimal".
# Assert on the MAGNITUDE of the median difference, not on its sign or on a bound
# taken from one draw -- both arms use common random numbers, but the cohort itself
# still varies with the solver thread count.
med <- ss_tbl$`Median AUC0-24,ss (mg*h/L)`
rel_diff <- 100 * abs(diff(med)) / mean(med)

# Realised ~25% here. The two models' typical clearances differ by 5.62 vs 4.37 L/h at
# the cohort median profile (a 25% exposure difference), which is a property of the
# published estimates, not of this draw. A 40% bound absorbs cohort noise while still
# going red if a parameter is mis-transcribed by a factor of 2 or a covariate term is
# dropped (either of which moves exposure by >= 80%).
stopifnot(rel_diff < 40)

# Both models must put the typical patient in a clinically plausible vancomycin range
# on this regimen -- an absolute sanity bound, not a race between the two arms.
stopifnot(all(med > 200), all(med < 900))
```

The two models place the median patient at different daily exposures on an identical
regimen -- the nonparametric model's higher typical clearance (5.62 vs 4.37 L/h at the
cohort median profile) translates into a correspondingly lower AUC. This is the
practical face of the Figure 2 finding that "on an average, Pmetrics estimated higher
values of CL and V than nlmixr2". The paper judged the clinical significance of these
distributional differences to be minimal, and reported near-identical *individualized*
(a posteriori) predictive performance for the two approaches -- nRMSE 13.8% versus
13.1%, MPE 0.16% versus 0.84% (Table 3) -- which is the metric that matters for MIPD,
where each patient's parameters are updated from their own measured levels.

# Model-fit statistics reported by the paper

For reference, these are the fit statistics from Table 3. They depend on the original
patient-level dataset, which is not publicly available, so they cannot be recomputed
here; they are reproduced only so the two packaged models can be read alongside the
performance that motivated them.

```{r table3-fit}
tibble::tribble(
  ~Metric,                          ~`Parametric (nlmixr2)`, ~`Nonparametric (Pmetrics)`,
  "r^2, population (a priori)",     0.597,                   0.531,
  "r^2, individual (a posteriori)", 0.896,                   0.917,
  "MPE %, population",              26.3,                    2.69,
  "MPE %, individual",              0.16,                    0.84,
  "nRMSE %, population",            27.2,                    30.7,
  "nRMSE %, individual",            13.8,                    13.1
) |>
  knitr::kable(caption = "Table 3 of Hughes 2024, model-fit rows (reproduced, not recomputed).")
```

# Assumptions and deviations

* **Infusion duration is an assumption.** The paper states that vancomycin was given by
  intermittent infusion and that peaks were drawn "within 2 h after the end of
  infusion", but never states the infusion duration. The simulations above infuse at
  1000 mg/h, the standard maximum vancomycin rate. This affects the shape of the peak
  but not AUC, and the closed-form gate (`AUC0-inf = Dose/CL`) is invariant to it.
* **Maintenance regimen is an assumption.** The protocol individualized maintenance
  dosing to an AUC target rather than fixing a dose. A flat 1000 mg q12h is used here so
  that the two models are compared on an identical regimen; it is not a dose the paper
  recommends.
* **Virtual-cohort distributional shapes are assumptions.** Table 1 reports medians and
  ranges only, so the distributions (normal for height and age, shifted-lognormal for
  BMI, lognormal for serum creatinine) were chosen to reproduce those medians and
  ranges. Race and ethnicity were not reported and are not simulated.
* **Eta correlations are not encoded.** The parametric model was fitted with a
  correlated `BLOCK(3)` on CL, Vc and Vp (SDC1 S3 declares
  `eta_CL + eta_V + eta_V2 ~ c(...)`), but the final off-diagonal estimates are
  published in neither the paper nor the supplement. Encoding them would mean inventing
  covariances, so both packaged models use independent etas. For scale, the NONMEM
  listing in SDC1 S4 implies correlations of 0.74 (CL-Vc), 0.12 (CL-Vp) and 0.66
  (Vc-Vp), but those belong to a different parameter set (see Errata) and are not
  transferable to the nlmixr2 column.
* **The nonparametric joint density is not recoverable.** NPAG's output is a discrete
  distribution over support points, and its shape -- including any multimodality, which
  is the main reason to run a nonparametric method at all -- cannot be reconstructed
  from the published summary table. The packaged nonparametric model approximates each
  marginal as an independent lognormal matching the reported median and %CV. The Table 2
  "Range" column spans the support points and is recorded in the parameter labels, but
  it is not an uncertainty interval and is not encoded as a variance. The Discussion
  itself notes that no subgroups could be identified in this analysis.
* **Pmetrics `lambda` is not encoded.** The nonparametric residual model is the assay
  polynomial `SD = C0 + C1*C` with `C0 = 0.5`, `C1 = 0.1` (encoded via `combined1()`,
  which is a linear sum, unlike nlmixr2's default quadrature combination). Pmetrics
  additionally fits an additive noise term, reported as `L = 0.1` in Table 3 and the
  SDC1 S5 `#Err` block, which nlmixr2 has no direct analogue for. Omitting it makes the
  encoded residual SD a very slight underestimate: at 10 mg/L, 1.500 encoded against
  1.503 in Pmetrics.
* **BSV scale conventions differ between the two files, deliberately.** For the
  parametric model the reported "%CV" is `100*sqrt(omega^2)`, which is demonstrated by
  the paper's own NONMEM `$OMEGA BLOCK(3)` reproducing the Table 2 NONMEM %CV row
  exactly (0.0473154, 0.0322236, 0.634656 -> 21.75 / 17.95 / 79.67 against a published
  21.8 / 18.0 / 79.7; the exact lognormal conversion would give 22.0 / 18.1 / 94.1).
  For the nonparametric model the %CV is a genuine descriptive coefficient of variation
  of the discrete distribution, so the exact lognormal conversion
  `omega^2 = log(CV^2 + 1)` is used instead. Both are documented at the point of use in
  the model files.
* **No published NCA to compare against.** Hughes 2024 reports no Cmax / Tmax / AUC /
  half-life table, so the NCA comparison above uses closed-form identities derived from
  the published parameters rather than a published NCA table.

## Errata and internal inconsistencies in the source

* **The parametric residual error is reported three different ways.** Table 2's nlmixr2
  column gives a proportional error of 16.8% with the additive row shown as "-";
  Table 3 gives "15.6% + 1.2 mg/L"; and the SDC1 S4 NONMEM listing gives
  `PROP = 0.157812` with `ADD = 0.001`. The 1.2 mg/L additive term in Table 3 appears
  nowhere else -- not in Table 2, not in the S3 nlmixr2 code (which declares
  `add_sd <- fixed(1e-3)`), not in S4. The packaged model uses Table 2's 16.8% with the
  additive fixed at 0.001 mg/L, because Table 2 is the paper's dedicated
  parameter-estimate table and its nlmixr2 column is the model being encoded, and
  because both supplement listings agree that the additive term is a nominal fixed
  value rather than an estimated 1.2 mg/L.
* **SDC1 S4 is labelled "NONMEM with final estimates" but does not match Table 2's
  NONMEM column.** S4 gives CL 5.21846, V1 65.1888, Q 6.65364, V2 62.6624,
  `THETA(8)` 0.859415, and an `$OMEGA BLOCK(3)` whose diagonal is
  0.0635929 / 0.0631089 / 0.520521 (25.2 / 25.1 / 72.2 %CV). Table 2's NONMEM column
  gives CL 5.06, Vc 64.34, Q 6.24, Vp 59.0, u3 0.851 and BSV 21.8 / 18.0 / 79.7 %CV.
  The Table 2 values are the ones reproduced by the Hughes control stream published in
  the Tong 2026 supplement (`modellib("Tong_2026_vancomycin_hughes")`: 5.0627, 64.339,
  6.2402, 59.019, 0.8509), so S4 appears to be an earlier or otherwise different run.
  Neither S4 nor the Table 2 NONMEM column is packaged from this paper; the NONMEM
  parameter set already exists in the library via the Tong 2026 extraction.
* **SDC1 S3 lists initial estimates, not final ones.** The nlmixr2 code block declares
  `tCL <- log(3)`, `tV <- log(60)`, `tQ <- log(2)`, `tV2 <- log(100)`,
  `tCRCL_CL <- 0.5` and unit-variance etas -- starting values, not the fitted results.
  It is used here only to establish model *structure* (which parameters are estimated,
  which are fixed, the residual-error form and the `BLOCK(3)` eta structure); every
  value comes from Table 2.
* **SDC1 S5 lists `TH1, 0.5` in the Pmetrics `#Pri` block**, which is a prior
  specification (a search range or starting value), not the fitted estimate. Table 2
  reports the fitted `u3` as 1.05 with 10% RSE, and that is what is encoded.
* **The Table 2 footnote's `u1` allometric exponent is not in either final model** --
  see the note under the Source trace table above.
