---
title: "Lamotrigine (Huo 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Lamotrigine (Huo 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Huo_2025_lamotrigine"))$reference`
* Description: `r rxode2::rxode(readModelDb("Huo_2025_lamotrigine"))$description`
* Article: <https://doi.org/10.2147/DDDT.S541597>

Huo 2025 is a two-centre retrospective therapeutic-drug-monitoring analysis of
lamotrigine (LTG) in Chinese women with epilepsy across pregnancy and the
postpartum period. Its purpose is dosing optimisation: apparent oral clearance
CL/F is the only structural parameter estimated, and everything else in the
paper -- the covariate model, the Monte Carlo simulations, and the Table 6
dosing grid -- follows from it.

# Population

128 women with epilepsy contributed 293 lamotrigine plasma concentrations,
collected at Beijing Tiantan Hospital and the Second Affiliated Hospital of
Zhejiang University between January 2015 and May 2024 (Huo 2025 Table 2). Mean
age was 28.24 +/- 3.79 years (range 19-36) and mean body weight 60.75 +/- 13.47
kg (range 40.00-114.00). All patients were on lamotrigine monotherapy at daily
doses of 25-800 mg given in one or two divided doses; 42.19% took at least one
concomitant antiseizure medication, of which only valproate (3.91%, n = 5)
entered the final model. Observations were distributed across the paper's five
peripregnancy stages as 13 / 63 / 111 / 84 / 22, and the postpartum samples
span 1-84 days after delivery.

The same information is available programmatically via
`readModelDb("Huo_2025_lamotrigine")()$population`.

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Huo_2025_lamotrigine.R` carries an in-file comment
naming its origin. They are collected here for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `d/dt(depot)`, `d/dt(central)`, `Cc` | n/a | Eqs 1-3, p 10246 |
| Exponential ("index") IIV, `theta_i = theta_TV * exp(eta_i)` | n/a | Eq 4, p 10246 |
| Combined error, `Cobs = Cpred * (1 + eps1) + eps2` | n/a | Eq 5, p 10246 |
| CL/F covariate model | n/a | Eq 6, p 10248 |
| `lka` (Ka) | 1.93 1/h, FIXED | Base Model p 10246; Table 4 Ka row (no %RSE, no CI) |
| `lvc` (V/F) | 68.8 L, FIXED | Eq 7 p 10249; Base Model p 10246; Table 4 Vd row "68.8 (fixed)" |
| `lcl` (CL/F) | 2.42 L/h | Table 4 Final Model CL row (%RSE 12.17, 95% CI 1.84-3.00) |
| `e_wt_cl` | 0.95 | Eq 6 exponent; Table 4 "BW on CL" (%RSE 22.98, 95% CI 0.52-1.38) |
| body-weight reference | 59.8 kg | Eq 6 denominator, p 10248 |
| `e_pregstage2_cl` | 0.28 | Coefficient block p 10249; Table 4 "Peripregnancy stage2 on CL" |
| `e_pregstage3_cl` | 0.59 | Coefficient block p 10249; Table 4 "Peripregnancy stage3 on CL" |
| `e_pregstage4_cl` | 0.57 | Coefficient block p 10249; Table 4 "Peripregnancy stage4 on CL" |
| `e_postpartum_cl` | -0.33 | Coefficient block p 10249; Table 4 "Peripregnancy stage5 on CL" |
| `e_conmed_vpa_cl` | -0.60 | Coefficient block p 10249; Table 4 "Inhibitors on CL" |
| stage boundaries (5, 14, 28 weeks GA; postpartum) | n/a | Table 1 row C ("Classification C was selected for the model") |
| `etalcl` | 0.10313 (= `log(0.3296^2 + 1)`) | Table 4 "IIV CL (CV%)" = 32.96%, via Eq 4 |
| `propSd` | 0.31 | Table 4 "sigma1 (multiplicative, CV)" (95% CI 0.27-0.34) |
| `addSd` | 0.004 mg/L | Table 4 "sigma2 (additive)" (95% CI 0.003-0.005) |

# Peripregnancy staging

The paper compared five candidate staging schemes (Table 1) and selected
Classification C, which splits pregnancy at 5, 14 and 28 gestational weeks and
adds a postpartum level. Classification C dropped the objective function by
181.38, against 169.33 / 125.60 / 112.67 / 49.17 for schemes A / D / B / E.

The packaged model derives the stage inside `model()` from two canonical
covariates -- `EGA` (maternal gestational age in weeks) and `TPP` (time
postpartum in weeks, 0 during pregnancy) -- so a user supplies ordinary
covariate columns rather than a pre-computed categorical.

```{r staging}
stages <- tibble::tribble(
  ~stage,               ~EGA, ~TPP, ~definition,
  "1 (< 5 wk GA)",         3,    0, "GA < 5 weeks (reference)",
  "2 (5-14 wk GA)",        9,    0, "5 <= GA < 14 weeks",
  "3 (14-28 wk GA)",      20,    0, "14 <= GA <= 28 weeks",
  "4 (> 28 wk to birth)", 34,    0, "28 < GA < delivery",
  "5 (postpartum)",        0,    4, "after delivery"
)
knitr::kable(stages, caption = "Huo 2025 Classification C (Table 1, row C).")
```

# Virtual cohort and simulation grid

Original observed data are not publicly available. The grid below is the
paper's own simulation design: the three representative body weights it used
for the Monte Carlo work (50, 65 and 80 kg), crossed with the five
peripregnancy stages and with valproate presence, and dosed at the daily
amounts Table 6 recommends for each cell.

```{r grid}
# Huo 2025 Table 6, "Dosage Recommendations Based on the Final Model",
# transcribed from the article PDF (the daily dose in mg for each
# stage / weight / inhibitor cell).
table6 <- tibble::tribble(
  ~stage_n, ~WT, ~CONMED_VPA, ~dose_mgd,
        1L,  50,          0L,      150,
        1L,  50,          1L,       50,
        1L,  65,          0L,      200,
        1L,  65,          1L,      100,
        1L,  80,          0L,      200,
        1L,  80,          1L,      150,
        2L,  50,          0L,      200,
        2L,  50,          1L,      100,
        2L,  65,          0L,      300,
        2L,  65,          1L,      100,
        2L,  80,          0L,      350,
        2L,  80,          1L,      150,
        3L,  50,          0L,      250,
        3L,  50,          1L,      150,
        3L,  65,          0L,      350,
        3L,  65,          1L,      200,
        3L,  80,          0L,      400,
        3L,  80,          1L,      250,
        4L,  50,          0L,      250,
        4L,  50,          1L,      150,
        4L,  65,          0L,      350,
        4L,  65,          1L,      200,
        4L,  80,          0L,      400,
        4L,  80,          1L,      250,
        5L,  50,          0L,      100,
        5L,  50,          1L,       50,
        5L,  65,          0L,      100,
        5L,  65,          1L,      100,
        5L,  80,          0L,      150,
        5L,  80,          1L,      150
) |>
  mutate(
    EGA  = c(3, 9, 20, 34, 0)[stage_n],
    TPP  = c(0, 0,  0,  0, 4)[stage_n],
    arm  = sprintf("S%d | %d kg | %s", stage_n, WT,
                   ifelse(CONMED_VPA == 1L, "VPA", "no VPA")),
    id   = row_number()
  )

stopifnot(nrow(table6) == 30L, !anyDuplicated(table6$arm))
```

The observation grid resolves the absorption peak (Tmax is near 2 h for
Ka = 1.93 1/h) and runs long enough for `aucinf.obs` to extrapolate reliably
even in the slowest arm (50 kg on valproate, terminal half-life about 43 h).

```{r events-single-dose}
obs_times <- sort(unique(c(
  seq(0, 12, by = 0.25),
  seq(13, 48, by = 1),
  seq(52, 336, by = 4)
)))

ev_dose <- table6 |>
  transmute(id, WT, EGA, TPP, CONMED_VPA, arm,
            time = 0, amt = dose_mgd, evid = 1L,
            cmt = "depot", Cc = NA_real_)

ev_obs <- table6 |>
  select(id, WT, EGA, TPP, CONMED_VPA, arm) |>
  tidyr::crossing(time = obs_times) |>
  mutate(amt = NA_real_, evid = 0L, cmt = "central", Cc = NA_real_)

events_sd <- bind_rows(ev_dose, ev_obs) |>
  arrange(id, time, desc(evid)) |>
  as.data.frame()

stopifnot(!anyDuplicated(unique(events_sd[, c("id", "time", "evid")])))
```

```{r simulate-single-dose}
mod <- readModelDb("Huo_2025_lamotrigine")

# Typical-value profiles: the paper's CL/F equation is a typical-value
# statement, so between-subject variability is switched off for the
# structural checks below.
mod_typ <- rxode2::zeroRe(mod)

sim_sd <- rxode2::rxSolve(
  mod_typ,
  events = events_sd,
  keep   = c("arm", "WT", "CONMED_VPA")
) |>
  as.data.frame()

stopifnot(nrow(sim_sd) > 0, all(sim_sd$Cc >= 0, na.rm = TRUE))
```

# PKNCA validation

`CL/F` is the paper's single estimated structural parameter, and for a
single dose it is recoverable exactly as `dose / AUC(0-inf)`. Running that
recovery through PKNCA turns Eq 6 into a falsifiable statement: a
mis-transcribed clearance, weight exponent, stage coefficient or valproate
coefficient moves the recovered `CL/F` for the affected arms and nothing else
in the vignette would notice.

```{r pknca}
sim_nca <- sim_sd |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

# Guarantee a time = 0 record per subject; for an extravascular dose the
# pre-dose concentration is 0.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, arm, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

dose_df <- events_sd |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, arm)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)

intervals <- data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE,
  cl.obs     = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(arm, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

stopifnot(nrow(nca_wide) == 30L, !anyNA(nca_wide$aucinf.obs))
```

## Recovered CL/F against Huo 2025 Equation 6

The reference column below is Eq 6 evaluated arithmetically from the values
printed in the paper -- it does not read anything out of the model object, so
the two sides of this comparison are genuinely independent.

```{r clf-compare}
# Huo 2025 Eq 6 (p 10248) plus the coefficient block on p 10249, written out
# from the printed numbers.
huo_eq6_clf <- function(WT, stage_n, CONMED_VPA) {
  stage_coef <- c(0, 0.28, 0.59, 0.57, -0.33)[stage_n]
  2.42 * (WT / 59.8)^0.95 * exp(stage_coef) * exp(-0.60 * CONMED_VPA)
}

published <- table6 |>
  transmute(
    arm,
    cl.obs = huo_eq6_clf(WT, stage_n, CONMED_VPA)
  )

cmp_clf <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_wide |> select(arm, cl.obs),
  reference = published,
  by        = "arm",
  units     = c(cl.obs = "L/h"),
  tolerance_pct = 5
)

knitr::kable(
  cmp_clf,
  caption = paste(
    "Apparent oral clearance recovered by PKNCA as dose / AUC(0-inf) from the",
    "packaged model, against Huo 2025 Equation 6 evaluated directly.",
    "* marks rows differing by more than 5%."
  ),
  align = c("l", "l", "r", "r", "r")
)
```

```{r clf-gate}
# This comparison is fully deterministic on both sides: typical-value
# parameters against closed-form arithmetic. The only difference is PKNCA's
# trapezoidal integration plus lambda-z extrapolation, so the bound is a
# numerical-accuracy bound, not a tolerance for model disagreement. Realised
# maximum across the 30 arms is well under 1%.
clf_chk <- nca_wide |>
  select(arm, cl_nca = cl.obs) |>
  left_join(published |> rename(cl_eq6 = cl.obs), by = "arm") |>
  mutate(pct_diff = 100 * (cl_nca - cl_eq6) / cl_eq6)

stopifnot(max(abs(clf_chk$pct_diff)) < 2)
cat(sprintf("Max |%% diff| in CL/F across the 30 arms: %.3f%%\n",
            max(abs(clf_chk$pct_diff))))
```

## Full NCA summary

```{r nca-table}
nca_wide |>
  left_join(table6 |> select(arm, dose_mgd), by = "arm") |>
  arrange(arm) |>
  mutate(across(c(cmax, tmax, aucinf.obs, half.life, cl.obs), ~ signif(.x, 3))) |>
  dplyr::rename(
    "Stage | weight | inhibitor" = arm,
    "Dose (mg)"                  = dose_mgd,
    "Cmax (mg/L)"                = cmax,
    "Tmax (h)"                   = tmax,
    "AUC0-inf (mg*h/L)"          = aucinf.obs,
    "t1/2 (h)"                   = half.life,
    "CL/F (L/h)"                 = cl.obs
  ) |>
  knitr::kable(
    caption = "Single-dose NCA of the typical-value profile for each Huo 2025 Table 6 cell.",
    align = c("l", "r", "r", "r", "r", "r", "r")
  )
```

# Published covariate effects

The paper states three quantitative consequences of Eq 6 in prose. Each is
checked below against the PKNCA-recovered clearances, which are model output
rather than restatements of the coefficients.

```{r covariate-effects}
clf <- nca_wide |>
  select(arm, cl.obs) |>
  left_join(table6 |> select(arm, stage_n, WT, CONMED_VPA), by = "arm")

# Stage fold-changes at matched weight and inhibitor status.
stage_ratio <- clf |>
  group_by(WT, CONMED_VPA) |>
  arrange(stage_n, .by_group = TRUE) |>
  mutate(ratio_vs_stage1 = cl.obs / cl.obs[stage_n == 1L]) |>
  ungroup() |>
  group_by(stage_n) |>
  summarise(model_pct = 100 * mean(ratio_vs_stage1), .groups = "drop") |>
  mutate(
    eq6_pct   = 100 * exp(c(0, 0.28, 0.59, 0.57, -0.33)),
    paper_pct = c(100, 131, 193, 199, 68)
  )

knitr::kable(
  stage_ratio |>
    mutate(across(c(model_pct, eq6_pct, paper_pct), ~ round(.x, 1))) |>
    dplyr::rename(
      "Peripregnancy stage"                = stage_n,
      "Packaged model (% of stage 1)"      = model_pct,
      "Eq 6 exponential (% of stage 1)"    = eq6_pct,
      "Huo 2025 Discussion (% of stage 1)" = paper_pct
    ),
  caption = "CL/F by peripregnancy stage, at matched body weight and valproate status."
)
```

```{r covariate-effect-gates}
# Deterministic: the packaged model must reproduce the exponential form of
# Eq 6 exactly (up to PKNCA integration error).
stopifnot(max(abs(stage_ratio$model_pct - stage_ratio$eq6_pct)) < 2)

# Valproate: Huo 2025 Discussion states "co-administration of VPA could
# decrease LTG CL/F by 46%". exp(-0.60) = 0.549, i.e. -45.1%.
vpa_ratio <- clf |>
  group_by(stage_n, WT) |>
  summarise(r = cl.obs[CONMED_VPA == 1L] / cl.obs[CONMED_VPA == 0L], .groups = "drop")
vpa_pct_drop <- 100 * (1 - mean(vpa_ratio$r))
stopifnot(abs(vpa_pct_drop - 46) < 2)
cat(sprintf("Valproate reduces CL/F by %.1f%% (paper: 46%%).\n", vpa_pct_drop))

# Body weight: Huo 2025 Discussion states "Patients >=90 kg showed 2.3-fold
# higher CL/F than those <=50 kg". The exponent alone gives (90/50)^0.95 =
# 1.75, so the stated 2.3-fold is a comparison of GROUP means rather than of
# the two thresholds; mean weights near 110 and 45 kg reproduce it exactly.
cat(sprintf("(90/50)^0.95 = %.2f-fold;  (110/45)^0.95 = %.2f-fold (paper: 2.3-fold).\n",
            (90 / 50)^0.95, (110 / 45)^0.95))
```

The three stage-effect columns agree to the digit for stages 2 and 5, but the
paper's Discussion percentages for stages 3 and 4 (193% and 199%) sit above
the exponentials of its own printed coefficients (180% and 177%). The gap is
explained by pregnancy weight gain: the Discussion percentages are
population-level CL/F ratios computed over the observations actually recorded
in each stage, and body weight enters CL/F alongside the stage term. About
4 kg of gain by stage 3 and 8 kg by stage 4, relative to stage 1, closes the
gap exactly.

```{r weight-gain-reconciliation}
wt_gain_needed <- tibble(
  stage_n   = c(2L, 3L, 4L),
  paper_pct = c(131, 193, 199),
  eq6_pct   = 100 * exp(c(0.28, 0.59, 0.57))
) |>
  mutate(
    # Solve (WT/57)^0.95 = paper_pct/eq6_pct for the extra weight, taking a
    # 57 kg stage-1 reference (the cohort mean of 60.75 kg includes the
    # later, heavier stages).
    implied_wt_kg   = 57 * (paper_pct / eq6_pct)^(1 / 0.95),
    implied_gain_kg = round(implied_wt_kg - 57, 1)
  )

knitr::kable(
  wt_gain_needed |>
    mutate(across(c(paper_pct, eq6_pct, implied_wt_kg), ~ round(.x, 1))) |>
    dplyr::rename(
      "Peripregnancy stage"          = stage_n,
      "Discussion (% of stage 1)"    = paper_pct,
      "Eq 6 alone (% of stage 1)"    = eq6_pct,
      "Implied mean weight (kg)"     = implied_wt_kg,
      "Implied gain vs stage 1 (kg)" = implied_gain_kg
    ),
  caption = paste(
    "Body-weight gain implied by the difference between the paper's",
    "population-level stage percentages and the exponentials of its own",
    "stage coefficients."
  )
)

# The reconciliation is only credible if the implied gains form an ordinary
# pregnancy weight trajectory. All three sides are deterministic arithmetic on
# printed numbers, so these bounds are exact rather than noise tolerances.
stopifnot(
  # Stage 2 (5-14 weeks GA) must need essentially no weight gain -- the
  # paper's 131% and Eq 6's 132% already agree, so anything more than about a
  # kilogram either way would mean the explanation does not hold there.
  abs(wt_gain_needed$implied_gain_kg[wt_gain_needed$stage_n == 2L]) < 1.5,
  # Stages 3 and 4 must need a positive gain, increasing with gestational age.
  wt_gain_needed$implied_gain_kg[wt_gain_needed$stage_n == 3L] > 1,
  wt_gain_needed$implied_gain_kg[wt_gain_needed$stage_n == 4L] >
    wt_gain_needed$implied_gain_kg[wt_gain_needed$stage_n == 3L],
  # ...and stay inside the range obstetric guidance describes for term
  # pregnancy, otherwise the gap would be too large for weight to explain.
  all(wt_gain_needed$implied_gain_kg <= 20)
)
```

# Reproducing Table 6: steady-state exposure under the recommended doses

Huo 2025 chose each Table 6 daily dose by Monte Carlo simulation against a
steady-state target range of 2.5-15 mg/L (Model-Informed LTG Dosing Regimens).
The paper does not state whether the daily dose was split, saying only that
lamotrigine was "typically given in one to two doses per day", so both
schedules are simulated.

```{r events-steady-state}
# 21 days of dosing: at least 11 half-lives even in the slowest arm.
make_ss_events <- function(n_per_day) {
  tau <- 24 / n_per_day
  dose_times <- seq(0, 21 * 24 - tau, by = tau)
  obs_grid   <- seq(20 * 24, 21 * 24, by = 0.5)

  d <- table6 |>
    select(id, WT, EGA, TPP, CONMED_VPA, arm, dose_mgd) |>
    tidyr::crossing(time = dose_times) |>
    mutate(amt = dose_mgd / n_per_day, evid = 1L, cmt = "depot") |>
    select(-dose_mgd)

  o <- table6 |>
    select(id, WT, EGA, TPP, CONMED_VPA, arm) |>
    tidyr::crossing(time = obs_grid) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")

  bind_rows(d, o) |> arrange(id, time, desc(evid)) |> as.data.frame()
}

ss_summary <- lapply(c(1L, 2L), function(npd) {
  ev <- make_ss_events(npd)
  rxode2::rxSolve(mod_typ, events = ev, keep = c("arm")) |>
    as.data.frame() |>
    dplyr::filter(!is.na(Cc), time >= 20 * 24) |>
    group_by(arm) |>
    summarise(
      ctrough = min(Cc),
      cav     = mean(Cc),
      cmax_ss = max(Cc),
      .groups = "drop"
    ) |>
    mutate(schedule = if (npd == 1L) "QD" else "BID")
}) |>
  bind_rows() |>
  left_join(table6 |> select(arm, stage_n, WT, CONMED_VPA, dose_mgd), by = "arm")
```

```{r table6-repro}
ss_summary |>
  select(arm, dose_mgd, schedule, ctrough, cav, cmax_ss) |>
  tidyr::pivot_wider(
    names_from  = schedule,
    values_from = c(ctrough, cav, cmax_ss)
  ) |>
  arrange(arm) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  select(arm, dose_mgd, ctrough_QD, ctrough_BID, cav_QD, cmax_ss_BID) |>
  dplyr::rename(
    "Stage | weight | inhibitor" = arm,
    "Table 6 dose (mg/day)"      = dose_mgd,
    "Ctrough, QD (mg/L)"         = ctrough_QD,
    "Ctrough, BID (mg/L)"        = ctrough_BID,
    "Cav (mg/L)"                 = cav_QD,
    "Cmax at SS, BID (mg/L)"     = cmax_ss_BID
  ) |>
  knitr::kable(
    caption = paste(
      "Typical-value steady-state exposure produced by the packaged model",
      "under each Huo 2025 Table 6 dosing recommendation.",
      "Cav is schedule-independent."
    ),
    align = c("l", rep("r", 5))
  )
```

```{r table6-gate}
cav <- ss_summary |> dplyr::filter(schedule == "QD") |> dplyr::pull(cav)

# Deterministic typical-value exposures, so a tight bound is appropriate.
# The paper's own dose-selection target is a steady-state concentration in
# 2.5-15 mg/L. Across all 30 cells the model puts Cav in a narrow band
# centred near 3 mg/L -- the signature of a dose table that was in fact
# generated by this clearance model. A mis-transcribed clearance, weight
# exponent, stage coefficient or dose would displace a subset of the cells
# and break the spread bound below.
cat(sprintf("Cav across the 30 Table 6 cells: median %.2f, range %.2f-%.2f mg/L\n",
            median(cav), min(cav), max(cav)))

stopifnot(
  # Centre of the distribution sits in the lower half of the paper's target
  # window, as a trough-targeted dose table must.
  median(cav) > 2.6, median(cav) < 3.6,
  # Coherence: the whole 30-cell table lands within a factor of 3.
  max(cav) / min(cav) < 3,
  # No cell is anywhere near the top of the therapeutic window.
  max(cav) < 15
)
```

```{r table6-in-range}
in_range <- ss_summary |>
  group_by(schedule) |>
  summarise(
    n_in_target = sum(ctrough >= 2.5 & ctrough <= 15),
    n_total     = dplyr::n(),
    median_ctrough = round(median(ctrough), 2),
    .groups = "drop"
  )

knitr::kable(
  in_range |>
    dplyr::rename(
      "Schedule"                       = schedule,
      "Cells with Ctrough in 2.5-15"   = n_in_target,
      "Cells"                          = n_total,
      "Median Ctrough (mg/L)"          = median_ctrough
    ),
  caption = "Typical-value troughs against the paper's 2.5-15 mg/L target."
)
```

The typical-value trough sits just below the 2.5 mg/L target floor in a
majority of cells, more so on a once-daily schedule than a twice-daily one.
This is a property of the paper's dose table rather than of the transcription:
Table 6 selects from a coarse grid in 50 mg steps, and the paper's own
diagnostics show where its concentrations actually lie -- Figure 2A plots
observed against predicted concentrations over 0-14 with the bulk between 1
and 4, and Figure 3's VPC has its median trough near 2-3. The packaged model
reproduces that placement. See the Errata below.

# Replicating Figure 5A: trough distributions in the 50 kg group

Figure 5A of Huo 2025 shows boxplots of simulated steady-state troughs for the
50 kg group at 50, 100, 150, 200, 250 and 300 mg/day, faceted by peripregnancy
stage. The version below uses the packaged model with its published
between-subject variability.

```{r figure-5a-cohort}
# set.seed() seeds R's RNG, not rxode2's. rxode2 partitions its streams per
# solver thread, so this cohort is reproducible here and different on a
# machine with a different thread count. Every assertion below is written to
# hold for any cohort the model can produce.
set.seed(20251209)

n_per_arm  <- 60L
fig5_doses <- c(50, 100, 150, 200, 250, 300)

fig5_arms <- tidyr::crossing(
  stage_n = 1:5,
  dose_mgd = fig5_doses
) |>
  mutate(
    EGA = c(3, 9, 20, 34, 0)[stage_n],
    TPP = c(0, 0, 0, 0, 4)[stage_n],
    arm_id = row_number()
  )

fig5_subj <- fig5_arms |>
  tidyr::crossing(rep = seq_len(n_per_arm)) |>
  mutate(
    id  = (arm_id - 1L) * n_per_arm + rep,
    WT  = 50,
    CONMED_VPA = 0L,
    stage_lbl = factor(
      c("Stage 1", "Stage 2", "Stage 3", "Stage 4", "Postpartum")[stage_n],
      levels = c("Stage 1", "Stage 2", "Stage 3", "Stage 4", "Postpartum")
    ),
    dose_lbl = factor(paste0(dose_mgd, " mg/d"),
                      levels = paste0(fig5_doses, " mg/d"))
  )

stopifnot(!anyDuplicated(fig5_subj$id))

fig5_dose_times <- seq(0, 21 * 24 - 12, by = 12)

fig5_ev <- bind_rows(
  fig5_subj |>
    tidyr::crossing(time = fig5_dose_times) |>
    mutate(amt = dose_mgd / 2, evid = 1L, cmt = "depot"),
  fig5_subj |>
    mutate(time = 21 * 24, amt = NA_real_, evid = 0L, cmt = "central")
) |>
  select(id, WT, EGA, TPP, CONMED_VPA, stage_lbl, dose_lbl, time, amt, evid, cmt) |>
  arrange(id, time, desc(evid)) |>
  as.data.frame()

stopifnot(!anyDuplicated(unique(fig5_ev[, c("id", "time", "evid")])))
```

```{r figure-5a}
sim_fig5 <- rxode2::rxSolve(
  mod, events = fig5_ev, keep = c("stage_lbl", "dose_lbl")
) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc))

stopifnot(nrow(sim_fig5) == nrow(fig5_subj))

ggplot(sim_fig5, aes(dose_lbl, Cc)) +
  geom_boxplot(outlier.size = 0.4) +
  geom_hline(yintercept = c(2.5, 15), linetype = "dashed", colour = "steelblue") +
  facet_wrap(~stage_lbl, nrow = 1) +
  scale_y_log10() +
  labs(
    x = "Daily dose", y = "Steady-state trough (mg/L)",
    title = "Simulated steady-state troughs, 50 kg group, twice-daily dosing",
    caption = paste(
      "Replicates Figure 5A of Huo 2025. Dashed lines mark the paper's",
      "2.5-15 mg/L target range."
    )
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r figure-5a-gates}
fig5_med <- sim_fig5 |>
  group_by(stage_lbl, dose_lbl) |>
  summarise(med = median(Cc), .groups = "drop")

# Cohort-derived, so assert trends and paper-stated absolute bounds only --
# never step-by-step ordering or an exact value from one draw.

# 1. Clearance rises through pregnancy and falls postpartum, so at a matched
#    dose the trough must fall from stage 1 to stage 4 and recover afterwards.
#    Stated as an END-TO-END trend, not adjacent-pair monotonicity.
med_by_stage <- fig5_med |>
  group_by(stage_lbl) |>
  summarise(med = median(med), .groups = "drop")
#    Both effects are large -- CL/F at stage 4 is 1.77 times its stage 1 value
#    and postpartum 0.72 times it -- so the bounds below are stated as
#    magnitudes with wide margins rather than as bare orderings. Pooled over
#    the six dose levels the ratios realised 0.43 and 1.29 at 1 solver thread.
stage_ratio_trough <- c(
  s4 = med_by_stage$med[med_by_stage$stage_lbl == "Stage 4"] /
       med_by_stage$med[med_by_stage$stage_lbl == "Stage 1"],
  pp = med_by_stage$med[med_by_stage$stage_lbl == "Postpartum"] /
       med_by_stage$med[med_by_stage$stage_lbl == "Stage 1"]
)
stopifnot(stage_ratio_trough[["s4"]] < 0.75, stage_ratio_trough[["pp"]] > 1.2)

# 2. Dose proportionality. The model is linear, so 300 mg/d must give exactly
#    six times the 50 mg/d trough -- but only for the SAME subject. Each arm
#    above draws its own 60 etas, so the ratio of arm medians is a ratio of two
#    independent sample medians and carries real noise: measured at 1 / 2 / 4
#    solver threads it realised 5.09-6.87 (max deviation from 6 of 0.91). The
#    exact statement therefore has to be made on typical values, and the
#    cohort version gets a bound outside the range observed across thread
#    counts. Do not tighten the second bound back toward the first.
sim_fig5_typ <- rxode2::rxSolve(mod_typ, events = fig5_ev,
                                keep = c("stage_lbl", "dose_lbl")) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc)) |>
  group_by(stage_lbl, dose_lbl) |>
  summarise(med = median(Cc), .groups = "drop")

dose_prop_typ <- sim_fig5_typ |>
  group_by(stage_lbl) |>
  summarise(ratio = med[dose_lbl == "300 mg/d"] / med[dose_lbl == "50 mg/d"],
            .groups = "drop")
stopifnot(max(abs(dose_prop_typ$ratio - 6)) < 1e-6)

dose_prop <- fig5_med |>
  group_by(stage_lbl) |>
  summarise(ratio = med[dose_lbl == "300 mg/d"] / med[dose_lbl == "50 mg/d"],
            .groups = "drop")
stopifnot(all(abs(dose_prop$ratio - 6) < 2))

# 3. Absolute placement against the paper's own diagnostics: Figure 4D plots
#    model-predicted concentrations on an axis running to 18 mg/L and
#    Figure 2A to 14 mg/L, with the bulk of both between 1 and 6. The 95th
#    percentile of the simulated cohort must sit inside that envelope. It
#    realised 7.18-7.23 at 1 / 2 / 4 solver threads, so the bound has ample
#    headroom while still failing on a unit or dose error of any size.
stopifnot(quantile(sim_fig5$Cc, 0.95) < 18)

cat(sprintf("Median trough by stage (50 kg, pooled over doses): %s mg/L\n",
            paste(sprintf("%.2f", med_by_stage$med), collapse = " / ")))
cat(sprintf("Stage 4 / stage 1 trough ratio: %.2f;  postpartum / stage 1: %.2f\n",
            stage_ratio_trough[["s4"]], stage_ratio_trough[["pp"]]))
cat(sprintf("Dose-proportionality ratio (300 vs 50 mg/d), typical value: %s\n",
            paste(sprintf("%.3f", dose_prop_typ$ratio), collapse = " / ")))
cat(sprintf("Dose-proportionality ratio (300 vs 50 mg/d), cohort medians: %s\n",
            paste(sprintf("%.2f", dose_prop$ratio), collapse = " / ")))
cat(sprintf("95th percentile of simulated troughs: %.2f mg/L\n",
            quantile(sim_fig5$Cc, 0.95)))
```

# Clearance trajectory across the peripregnancy period

```{r clearance-trajectory}
traj <- tibble(EGA = c(seq(0, 40, by = 0.5))) |>
  mutate(
    TPP = 0,
    phase = "Pregnancy"
  ) |>
  bind_rows(
    tibble(EGA = 0, TPP = seq(0.5, 12, by = 0.5), phase = "Postpartum")
  ) |>
  mutate(
    x = ifelse(phase == "Pregnancy", EGA, 40 + TPP),
    stage_coef = dplyr::case_when(
      TPP > 0   ~ -0.33,
      EGA <  5  ~  0,
      EGA < 14  ~  0.28,
      EGA <= 28 ~  0.59,
      TRUE      ~  0.57
    ),
    clf = 2.42 * (60.75 / 59.8)^0.95 * exp(stage_coef)
  )

ggplot(traj, aes(x, clf, colour = phase)) +
  geom_step(linewidth = 0.9) +
  geom_vline(xintercept = c(5, 14, 28, 40), linetype = "dotted") +
  labs(
    x = "Gestational age (weeks), then weeks postpartum after 40",
    y = "CL/F (L/h) at the cohort mean weight of 60.75 kg",
    colour = NULL,
    title = "Lamotrigine apparent clearance across the peripregnancy period",
    caption = paste(
      "Replicates the clearance panel of Figure 5 of Huo 2025. Dotted lines",
      "mark the Classification C nodes at 5, 14 and 28 weeks and delivery."
    )
  )
```

# Assumptions and deviations

## Errata in the source

- **V/F is 68.8 L, not 59.8 L.** The Discussion (p 10251) says "this study set
  the Vd value as 59.8 L". That conflicts with three other statements in the
  same paper: the Base Model paragraph ("Ka and V/F were fixed at 1.93 h-1 and
  68.8 L"), Eq 7 ("Vd(L) = 68.8 fixed"), and Table 4, whose Vd row reads
  "68.8 (fixed)" in all three of its Base Model, Final Model and Bootstrap
  columns. 59.8 is the body-weight normalisation constant of Eq 6, so the
  Discussion sentence is a transcription slip in which the weight constant was
  written where the volume belonged. The model uses 68.8 L.

- **Concentrations are in mg/L; the "mmol/L" labels are wrong.** Table 4 labels
  its additive residual-error row "sigma2 (additive, mmol/L)", and the axes of
  Figure 2 ("Observed(mmol/L)", "Population predicted concentration(mmol/L)")
  and Figure 3 ("Observed (mmol/L)") carry the same unit. The paper's own
  Figure 4D contradicts them: it plots the same quantity as "Predicted
  concentrations (mg/L)" over an identical 0-18 range. mg/L is also the only
  reading consistent with the rest of the article -- the assay is linear from
  1.37 to 20.9 mg/L with an LLOQ of 1.37 mg/L, and the therapeutic target is
  2.5-15 mg/L. Lamotrigine's molar mass of 256.09 g/mol puts 3 mg/L at
  0.012 mmol/L, three orders of magnitude away from the plotted values. The
  model therefore carries `addSd` in mg/L.

- **The additive residual error is negligible as reported.** Table 4 gives
  sigma2 = 0.004 in the final model, down from 0.38 in the base model, with a
  95% CI of 0.003-0.005 and a bootstrap 95% CI of 0.003729-0.003733 that is far
  tighter than any other row in the table. Against an LLOQ of 1.37 mg/L an
  additive SD of 0.004 mg/L contributes nothing, so the error model is
  effectively proportional. The value is transcribed as printed and not
  adjusted.

- **Table 4's "Inhibitors on CL" bootstrap 95% CI is printed as
  (-1.09, -1.28)**, which is inverted -- the lower bound exceeds the upper. The
  point estimate is unaffected and only the point estimate is used here.

- **The Discussion's stage percentages exceed the exponentials of its own
  coefficients for stages 3 and 4.** The Discussion reports CL/F at 131%, 193%,
  199% and 68% of stage 1, while `exp()` of the printed stage coefficients
  gives 132%, 180%, 177% and 72%. Stages 2 and 5 agree; stages 3 and 4 do not.
  The reconciliation shown above is pregnancy weight gain: the percentages are
  population-level ratios over the observations recorded in each stage, and
  body weight enters CL/F alongside the stage term with an exponent of 0.95.
  About 4 kg by stage 3 and 8 kg by stage 4 reproduces the published
  percentages exactly, which is an ordinary weight trajectory. The packaged
  model encodes the printed equation, per the standing rule that a printed
  equation outranks a prose restatement.

- **The Discussion's dose-adjustment bands are looser than Table 6.** The text
  says the postpartum daily dose should be "reduced by 0-25%" relative to
  stage 1, but Table 6's postpartum cells are 100 / 100 / 150 mg/day against
  stage 1's 150 / 200 / 200, i.e. reductions of 33% / 50% / 25%. Similarly the
  stage 2 band is given as "increased by 25-50%" while the 80 kg cell rises
  from 200 to 350 mg/day (+75%). Table 6 is the model's actual output and is
  used here; the prose bands are approximations.

## Modelling assumptions

- **Peripregnancy stage is derived inside `model()` from `EGA` and `TPP`.** The
  paper fits a five-level categorical, but every level boundary is an explicit
  gestational-week cutoff printed in Table 1 row C, so the banding is part of
  the published model rather than a data-assembly convention. Carrying it on
  the two ratified canonical covariates -- `EGA` for maternal gestational age
  and `TPP` for time since delivery -- follows the `EGA` register entry, which
  directs gestational-week stratification onto `EGA` rather than a new
  trimester-indicator canonical. Set `TPP = 0` for all pregnancy records; any
  positive `TPP` selects the postpartum level, because the paper models
  postpartum as a single step and not as a continuous recovery.

- **`CONMED_VPA` rather than `CONMED_UGT_INH`.** The paper names the covariate
  "inhibitor", but its Data Collection section defines the enzyme-inhibitor
  class as containing exactly one drug, valproic acid, and both the Discussion
  and Table 6 refer to VPA by name. The drug-specific canonical is therefore
  correct. `exp(-0.60) = 0.549` reproduces the paper's stated 46% reduction, and
  matches the independent estimate of `exp(-0.61)` in
  `Damnjanovic_2023_lamotrigine`, a paediatric lamotrigine model with the same
  covariate.

- **IIV is carried as a variance derived from the reported CV%.** Table 4
  reports IIV on CL/F as a CV of 32.96%. Eq 4 is the exponential random-effect
  model, so `omega^2 = log(CV^2 + 1) = 0.10313`. Reading the 32.96% as the
  log-scale SD directly would give 0.10864 instead, a 5% difference in variance
  and 2.6% in SD; the log-normal identity is used because that is the form Eq 4
  specifies. No other parameter carries IIV, because Ka and V/F were fixed --
  a limitation the paper itself flags.

- **The dosing schedule behind Table 6 is unstated.** The paper reports daily
  doses and says lamotrigine was "typically given in one to two doses per day"
  without saying which split the Monte Carlo simulations used, so both
  schedules are shown above rather than one being assumed.

- **Body weight is treated as time-varying.** The paper does not state whether
  a per-record or a baseline weight was used. The per-record reading is adopted
  because it is what reconciles the Discussion's stage percentages with Eq 6,
  as shown above.

- **The 59.8 kg normalisation constant is not identified in the paper.** It is
  not the cohort mean weight, which Table 2 gives as 60.75 kg; it is most
  plausibly the cohort median. It is transcribed exactly as printed in Eq 6.

- **No supplement was needed.** The paper's Supplementary Material holds a
  covariate correlation matrix (Figure S1), the forward-addition screen
  (Table S1), and comparisons with three earlier peripregnancy lamotrigine
  models (Tables S2 and S3). None of these carries a final-model parameter
  value: every value in the packaged model comes from the main article's
  Eqs 6-7, its p 10249 coefficient block, and Table 4.

## Screened but not retained

Huo 2025 evaluated age, BSA, BMI, albumin, AST, ALT, total protein, total and
direct bilirubin, blood urea nitrogen, creatinine, and the enzyme-inducing
comedications carbamazepine, oxcarbazepine and phenobarbital, and retained none
of them. These are recorded in the model file's `covariatesDataExcluded` list
with the paper's stated reason in each case -- a narrow 19-36 year age span,
liver and kidney function within normal limits for almost every participant,
collinearity with body weight, or too few exposed patients (3 on carbamazepine,
5 on phenobarbital). Serum total protein has no canonical covariate column and
so is documented here rather than in the model file.
