---
title: "Midazolam N-glucuronidation in vitro (Hyland 2009)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Midazolam N-glucuronidation in vitro (Hyland 2009)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(ggplot2)
```

# Model and source

Hyland and colleagues asked whether the direct N-glucuronidation of midazolam,
already known from human liver microsome (HLM) incubations, happens in people.
They isolated an authentic midazolam N-glucuronide (MDZG) standard, confirmed by
NMR that the glucuronic acid sits on the alpha-nitrogen of the imidazole ring,
quantified MDZG in the urine of six healthy men, screened twelve recombinant
UGTs to find which enzyme is responsible, and characterised the enzyme kinetics
of the pathway.

The fitted model is that last piece: MDZG formation follows neither
Michaelis-Menten kinetics nor a simple Hill equation. The Eadie-Hofstee plots
(source Figure 2) bend both ways, showing allosteric activation at low substrate
*and* substrate inhibition at high substrate, so the authors fitted a two-site
substrate-inhibitory cooperative binding model (source Equation 2) separately to
two enzyme systems.

Those are two independent curve fits on two different enzyme preparations, so
they are two model files.

```{r models}
mod_hlm <- rxode2::rxode(readModelDb("Hyland_2009_midazolam_hlm"))
mod_ugt <- rxode2::rxode(readModelDb("Hyland_2009_midazolam_rugt1a4"))
```

* Citation: `r mod_hlm$reference`
* Article: <https://doi.org/10.1111/j.1365-2125.2009.03386.x>
* PubMed Central open-access copy: <https://pmc.ncbi.nlm.nih.gov/articles/PMC2679108/>

| Model file | Enzyme system | Source |
|---|---|---|
| `Hyland_2009_midazolam_hlm` | Pooled human liver microsomes (HLM101) | Table 4, column "MDZ HLM"; Figure 3 open circles |
| `Hyland_2009_midazolam_rugt1a4` | Recombinant human UGT1A4 Supersomes | Table 4, column "rUGT1A4"; Figure 3 closed circles |

UGT1A4 was the only enzyme of the twelve recombinant UGTs screened (1A1, 1A3,
1A4, 1A6, 1A7, 1A8, 1A9, 1A10, 2B4, 2B7, 2B15, 2B17, alongside a control) that
conjugates midazolam directly, favoured over UGT2B4 and UGT2B7 by more than
100-fold and 400-fold respectively (Results, "Initial screening"). The two
files therefore describe the same enzyme in two preparations, which is why their
parameters are close but not identical.

## Experimental system

```{r population}
pop <- mod_hlm$population
fields <- c("system", "matrix", "initiation", "temperature", "incubation_time",
            "substrate_range", "replicates", "quantification")
stopifnot(all(fields %in% names(pop)))   # fail loudly rather than render an empty table
knitr::kable(
  data.frame(
    Field = fields,
    Value = vapply(pop[fields], function(z) paste(z, collapse = "; "), character(1))
  ),
  row.names = FALSE,
  caption = "Incubation conditions (Materials and methods, 'Microsomal incubations')."
)
```

This is an initial-rate assay, so the model holds the substrate concentration
constant and lets the glucuronide accumulate:

* `midazolam` -- the static substrate concentration in the incubation (umol/L).
  The "dose" is the concentration placed in the tube at time 0.
* `mdzg` -- glucuronide formed, in pmol per mg of microsomal protein.
* `Cc` -- the observable, which here is the instantaneous formation rate `v` in
  pmol/min/mg protein. This is what source Figure 3 plots, and `Cc` is the
  library's required single-output observation name.

# Source trace

Every value in both `ini()` blocks, and every non-trivial line of `model()`.

```{r source-trace}
trace <- tibble::tribble(
  ~Quantity,   ~HLM,    ~rUGT1A4, ~Source,
  "Vmax (pmol/min/mg protein)", "445 +/- 1",     "427 +/- 0.5",   "Table 4",
  "Ks (umol/L)",                "46 +/- 5",      "64 +/- 6.4",    "Table 4 (called 'Km' in the Abstract)",
  "Ki (umol/L)",                "58 +/- 6",      "79 +/- 8",      "Table 4",
  "alpha (unitless)",           "18.3 +/- 0.0007", "14.9 +/- 0.0008", "Table 4",
  "n (unitless)",               "2.3 +/- 0.05",  "2.6 +/- 0.07",  "Table 4",
  "beta (unitless)",            "0.01, FIXED",   "0.01, FIXED",   "Results, 'MDZG enzyme kinetics'",
  "R-squared of the fit",       "0.962",         "0.986",         "Table 4",
  "Rate law",                   "Equation 2",    "Equation 2",    "Results, p. 449 (see deviation below)",
  "Residual error",             "not reported",  "not reported",  "Table 4 gives R-squared only"
)
knitr::kable(trace, caption = "Source trace for both model files.")
```

## Units

Dimensional analysis is not optional for a mechanistic model, and here it is
what identifies a defect in the printed equation (next section).

| Symbol | Units | Role |
|---|---|---|
| `midazolam` | umol/L | substrate concentration `[S]`, held constant |
| `mdzg` | pmol/mg protein | accumulated product |
| `vmax` | pmol/min/mg protein | catalytic rate at full saturation |
| `ks`, `ki` | umol/L | dissociation constants |
| `alpha`, `beta`, `nhill` | unitless | shape factors |
| `v` = `Cc` | pmol/min/mg protein | formation rate |
| `d/dt(mdzg)` = `v` | (pmol/mg protein) / min | product accumulation |

# The equation as typeset does not reproduce the paper

Source Equation 2 is printed as

$$v = \frac{V_{max} \times \left[\dfrac{1}{K_s} + \dfrac{\beta \times [S]^n}{\alpha K_s K_i}\right]}{\dfrac{1}{[S]^n} + \dfrac{1}{K_s} + \dfrac{1}{K_i} + \dfrac{[S]^n}{\alpha K_s K_i}}$$

with `Ks` and `Ki` carrying no exponent. Both model files instead use the form
with $K_s^n$ and $K_i^n$. Four independent checks show that is what was actually
fitted; three are below and the fourth (the range of the Eadie-Hofstee axis)
falls out of the initial-rate calculation further down.

**(a) The printed form is dimensionally inconsistent.** Its denominator adds
$1/[S]^n$, in $\mu M^{-n}$ with $n = 2.3$, to $1/K_s$, in $\mu M^{-1}$. Those
are commensurable only when $n = 1$. Raising `Ks` and `Ki` to the $n$-th power
-- the standard Hill generalisation of a dissociation constant -- is the unique
assignment that makes every term $\mu M^{-n}$. It is also exactly the source
the Methods cite: "a modification of equation 11 from Shou *et al.*" combined
with the Hill equation, i.e. the Shou two-site substrate-inhibition model with
$[S]$ replaced by $[S]^n$.

**(b) The printed form does not reproduce the paper's own Figure 3.**

```{r falsify, fig.height = 4.5}
# The rate law under both readings, evaluated directly from the Table 4 values.
v_printed <- function(S, Vmax, Ks, Ki, a, n, b = 0.01) {
  u <- S^n
  Vmax * (1 / Ks + b * u / (a * Ks * Ki)) /
    (1 / u + 1 / Ks + 1 / Ki + u / (a * Ks * Ki))
}
v_hill <- function(S, Vmax, Ks, Ki, a, n, b = 0.01) {
  u <- S^n; A <- Ks^n; B <- Ki^n; C <- a * A * B
  Vmax * (u / A + b * u * u / C) / (1 + u / A + u / B + u * u / C)
}

pars <- list(HLM = c(445, 46, 58, 18.3, 2.3), rUGT1A4 = c(427, 64, 79, 14.9, 2.6))
Sgrid <- seq(1, 1000, by = 1)
cmp <- bind_rows(lapply(names(pars), function(nm) {
  p <- pars[[nm]]
  bind_rows(
    data.frame(system = nm, reading = "as typeset",
               S = Sgrid, v = do.call(v_printed, c(list(Sgrid), as.list(p)))),
    data.frame(system = nm, reading = "with Ks^n, Ki^n",
               S = Sgrid, v = do.call(v_hill, c(list(Sgrid), as.list(p))))
  )
}))

ggplot(cmp, aes(S, v, colour = reading)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~system) +
  labs(x = "Midazolam (umol/L)", y = "v (pmol/min/mg protein)",
       colour = NULL,
       title = "Two readings of Equation 2",
       subtitle = "Figure 3 of the paper peaks near 100 umol/L at about 230 (HLM) and 217 (rUGT1A4)") +
  theme_bw() + theme(legend.position = "bottom")
```

Source Figure 3 plots both fitted curves against midazolam concentration on a
0-1000 umol/L axis with the rate axis labelled every 20 units to 240. Two facts
are readable off that axis without any digitisation: **the maximum of each
fitted curve lies to the left of the 200 umol/L tick, and it lies between the
220 and 240 rate labels.** The reading with $K_s^n$ and $K_i^n$ satisfies both;
the form as typeset puts the maximum an order of magnitude too far left.

```{r falsify-gate}
peak <- function(f, p) {
  o <- optimize(function(s) do.call(f, c(list(s), as.list(p))), c(0.1, 1000), maximum = TRUE)
  c(S = o$maximum, v = o$objective)
}
pk <- bind_rows(lapply(names(pars), function(nm) {
  data.frame(system = nm,
             reading = c("as typeset", "with Ks^n, Ki^n"),
             rbind(peak(v_printed, pars[[nm]]), peak(v_hill, pars[[nm]])))
}))
knitr::kable(pk, digits = 1,
             caption = "Location and height of the fitted maximum under each reading.")

# Bounds read off the printed axis of Figure 3, deliberately generous:
# the peak sits well inside the first 200 umol/L division and between the
# 220 and 240 rate labels. These are DETERMINISTIC quantities -- the models
# carry no random effects -- so a tight bound is correct here.
ok <- pk$reading == "with Ks^n, Ki^n"
stopifnot(
  all(pk$S[ok]  > 60  & pk$S[ok]  < 160),
  all(pk$v[ok]  > 205 & pk$v[ok]  < 245),
  # ... and the form as typeset is excluded by the same bounds, by a wide margin.
  all(pk$S[!ok] < 20)
)
```

**(c) The printed form contradicts source Figure 4.** The inhibition experiment
incubated midazolam "at a concentration approximating to the `Km`" and reports
uninhibited control rates of 160.2, 154.2 and 150.0 pmol/mg/min for the
ketoconazole, itraconazole and trifluoperazine runs. The Abstract gives that
`Km` as 46 umol/L for HLM.

```{r fig4-gate}
ctrl <- c(KTZ = 160.2, ITZ = 154.2, TFP = 150.0)   # Figure 4 legend
at_km <- c("as typeset"      = v_printed(46, 445, 46, 58, 18.3, 2.3),
           "with Ks^n, Ki^n" = v_hill(46, 445, 46, 58, 18.3, 2.3))
knitr::kable(
  data.frame(Reading = names(at_km),
             `Predicted rate at 46 umol/L` = round(at_km, 1),
             `vs Figure 4 controls (150.0-160.2)` =
               sprintf("%+.0f%% to %+.0f%%",
                       100 * (at_km / max(ctrl) - 1), 100 * (at_km / min(ctrl) - 1)),
             check.names = FALSE),
  row.names = FALSE,
  caption = "Predicted uninhibited rate at the reported Km against the Figure 4 controls."
)
stopifnot(
  # The Hill reading lands within 20% of the observed control window;
  # the form as typeset is low by a factor of about 2.6.
  abs(at_km[["with Ks^n, Ki^n"]] / mean(ctrl) - 1) < 0.20,
  at_km[["as typeset"]] / mean(ctrl) < 0.5
)
```

This is a defect in the typesetting of Equation 2, not a modelling choice: the
exponents on `Ks` and `Ki` were lost. The model files reproduce the fit the
paper actually shows.

# Replicating Figure 3

The models are solved as the experiment was run -- a 20 minute incubation at a
fixed midazolam concentration -- and the formation rate is read off at the end.

```{r solve}
solve_curve <- function(mod, S, minutes = 20) {
  ev <- do.call(rbind, lapply(seq_along(S), function(i) {
    data.frame(id = i, time = c(0, minutes), amt = c(S[i], NA_real_),
               evid = c(1L, 0L), cmt = "midazolam")
  }))
  out <- rxode2::rxSolve(mod, ev, returnType = "data.frame")
  out <- out[out$time == minutes, ]
  out$S <- S[out$id]
  out
}

Sobs <- c(seq(1, 300, by = 1), seq(305, 1000, by = 5))
curve_hlm <- solve_curve(mod_hlm, Sobs)
curve_ugt <- solve_curve(mod_ugt, Sobs)
fig3 <- bind_rows(
  transform(curve_hlm, system = "HLM101"),
  transform(curve_ugt, system = "rUGT1A4")
)
```

```{r fig3, fig.height = 4.5}
ggplot(fig3, aes(S, Cc, colour = system)) +
  geom_line(linewidth = 0.9) +
  scale_x_continuous(breaks = seq(0, 1000, by = 200), limits = c(0, 1000)) +
  scale_y_continuous(breaks = seq(0, 240, by = 20), limits = c(0, 245)) +
  labs(x = "Midazolam (umol/L)", y = "V (pmol/min/mg)", colour = NULL,
       title = "Replicates Figure 3 of Hyland 2009",
       subtitle = "MDZG kinetics fitted to a two-site substrate-inhibition cooperative binding model") +
  theme_bw() + theme(legend.position = "bottom")
```

The published curves rise steeply to a maximum near 100-120 umol/L and then decay
across the rest of the axis, running close together throughout with the HLM curve
generally the upper of the two. That is reproduced, with one caveat recorded
below: between roughly 150 and 345 umol/L the implemented curves cross, putting
rUGT1A4 up to 3% above HLM where the published figure keeps HLM on top. The
offset is a few pmol/min/mg and is consistent with Table 4 reporting `alpha` and
`n` -- the two parameters that most strongly set the curve shape -- to only two
or three significant figures. It is recorded rather than tuned away.

```{r fig3-gate}
pk_hlm <- curve_hlm[which.max(curve_hlm$Cc), ]
pk_ugt <- curve_ugt[which.max(curve_ugt$Cc), ]
at <- function(d, s) d$Cc[which.min(abs(d$S - s))]

fig3_tab <- data.frame(
  Feature = c("HLM peak rate (pmol/min/mg)", "HLM peak location (umol/L)",
              "rUGT1A4 peak rate (pmol/min/mg)", "rUGT1A4 peak location (umol/L)",
              "HLM rate at 200 umol/L", "HLM rate at 400 umol/L",
              "HLM rate at 800 umol/L", "HLM vs rUGT1A4 at 800 umol/L"),
  Model = c(sprintf("%.1f", c(pk_hlm$Cc, pk_hlm$S, pk_ugt$Cc, pk_ugt$S,
                              at(curve_hlm, 200), at(curve_hlm, 400), at(curve_hlm, 800))),
            sprintf("%+.1f%%", 100 * (at(curve_hlm, 800) / at(curve_ugt, 800) - 1))),
  `Read from Figure 3` = c("between the 220 and 240 labels", "left of the 200 tick",
                           "between the 200 and 220 labels", "left of the 200 tick",
                           "just under 180", "between 60 and 90",
                           "between 10 and 30", "HLM above rUGT1A4 in the tail"),
  check.names = FALSE
)
knitr::kable(fig3_tab, caption = "Model against features readable off the printed axis of Figure 3.")

stopifnot(
  pk_hlm$Cc > 220 & pk_hlm$Cc < 240,
  pk_ugt$Cc > 200 & pk_ugt$Cc < 220,
  pk_hlm$S  > 60  & pk_hlm$S  < 200,
  pk_ugt$S  > 60  & pk_ugt$S  < 200,
  at(curve_hlm, 200) > 160 & at(curve_hlm, 200) < 190,
  at(curve_hlm, 400) > 60  & at(curve_hlm, 400) < 90,
  at(curve_hlm, 800) > 10  & at(curve_hlm, 800) < 30,
  # HLM sits above rUGT1A4 in the tail, where the published gap is real
  # (13-20%). Deliberately NOT asserted at 200-300 umol/L: there the two
  # published curves are nearly coincident and the implemented pair cross,
  # so an ordering test would be a knife-edge on Table 4's rounding rather
  # than a check on the model. See the caveat above and the Errata.
  at(curve_hlm, 600) > at(curve_ugt, 600),
  at(curve_hlm, 800) > at(curve_ugt, 800)
)

# The published curves run close together across the descending limb; so do
# these. Bound the separation rather than its sign where the sign is not robust.
sep <- curve_hlm %>%
  select(S, hlm = Cc) %>%
  inner_join(select(curve_ugt, S, ugt = Cc), by = "S") %>%
  filter(S >= 150, S <= 800) %>%
  mutate(pct = 100 * (hlm / ugt - 1))
cat(sprintf("HLM vs rUGT1A4 over 150-800 umol/L: %.1f%% to %+.1f%%\n",
            min(sep$pct), max(sep$pct)))
stopifnot(min(sep$pct) > -6, max(sep$pct) < 25)
```

# Replicating Figure 2 (Eadie-Hofstee)

Source Figure 2 plots `v` against `v/[S]`. A Michaelis-Menten enzyme gives a
straight line with slope `-Km`; the published plots are hooked in both
directions, which is the diagnostic the authors used to reject
Michaelis-Menten. The model reproduces that shape, and the reason it is hooked
rather than straight is worth stating: the low-substrate limb bends because of
the Hill exponent (allosteric activation) and the high-substrate limb bends back
because of substrate inhibition.

```{r fig2, fig.height = 4.5}
eh <- fig3 %>% filter(S >= 10, S <= 800) %>% mutate(v_over_S = Cc / S)
ggplot(eh, aes(v_over_S, Cc, colour = system)) +
  geom_path(linewidth = 0.9) +
  labs(x = "v/[S]", y = "v (pmol/min/mg)", colour = NULL,
       title = "Replicates Figure 2 of Hyland 2009",
       subtitle = "Eadie-Hofstee plots over the studied 10-800 umol/L range") +
  theme_bw() + theme(legend.position = "bottom")
```

```{r fig2-gate}
# The signature of the published plot: v/[S] is NOT monotone in [S] -- it rises
# with the allosteric limb and falls again -- so the curve doubles back. A
# Michaelis-Menten enzyme could not produce this.
hook <- eh %>% group_by(system) %>%
  summarise(max_v_over_S = max(v_over_S),
            v_over_S_at_lowest_S = v_over_S[which.min(S)],
            v_over_S_at_highest_S = v_over_S[which.max(S)], .groups = "drop")
knitr::kable(hook, digits = 3,
             caption = "v/[S] doubles back: its maximum lies strictly inside the studied range.")
stopifnot(
  # Interior maximum in v/[S] => the Eadie-Hofstee curve is hooked, not linear.
  all(hook$max_v_over_S > hook$v_over_S_at_lowest_S),
  all(hook$max_v_over_S > hook$v_over_S_at_highest_S),
  # Figure 2's x axis runs to 4; both systems stay inside it.
  all(hook$max_v_over_S < 4)
)
```

# Mechanistic checks

## Product accumulation

The substrate is held constant, so the accumulated glucuronide must be exactly
the rate times the incubation time. This is a pure numerical-identity check --
both sides come from the same solve -- so it is asserted tightly.

```{r accumulation}
acc <- solve_curve(mod_hlm, c(10, 46, 100, 200, 400, 800), minutes = 20)
knitr::kable(
  data.frame(`Midazolam (umol/L)` = acc$S,
             `v (pmol/min/mg)` = round(acc$Cc, 2),
             `MDZG at 20 min (pmol/mg)` = round(acc$mdzg, 1),
             `v x 20 min` = round(acc$Cc * 20, 1),
             check.names = FALSE),
  row.names = FALSE,
  caption = "Accumulated MDZG equals the formation rate times the incubation time."
)
stopifnot(isTRUE(all.equal(acc$mdzg, acc$Cc * 20, tolerance = 1e-6)))
```

## The initial-rate assumption

Holding the substrate constant is only legitimate if the incubation consumes a
small fraction of it. Over 20 minutes at 0.5 mg protein/mL the amount turned
over per mL is `v * 20 * 0.5` pmol, so the fraction of substrate consumed is

$$\frac{v \times 20\,\text{min} \times 0.5\,\text{mg/mL}}{[S]} = 0.01 \times \frac{v}{[S]}$$

which is 1/100th of the Eadie-Hofstee ordinate. Source Figure 2 bounds `v/[S]`
below 4 across the studied range, so at most about 4% of the substrate is
consumed -- small enough for the initial-rate treatment the published rate law
assumes, but worth stating rather than waving at.

```{r depletion}
protein_mg_per_mL <- 0.5    # Methods, 'MDZG kinetics': HLM at 0.5 mg protein/mL
minutes <- 20               # Methods, 'MDZG kinetics': 20 min incubation

# v [pmol/(min*mg)] * t [min] * p [mg/mL] = pmol/mL; 1 pmol/mL = 1e-3 umol/L.
depletion <- fig3 %>%
  filter(S >= 10, S <= 800) %>%
  mutate(pct_consumed = 100 * Cc * minutes * protein_mg_per_mL * 1e-3 / S)

worst <- depletion %>% slice_max(pct_consumed, n = 1)
shown <- depletion %>%
  filter(system == "HLM101", S %in% c(10, 46, 100, 400, 800)) %>%
  select(`Midazolam (umol/L)` = S, `v (pmol/min/mg)` = Cc,
         `Substrate consumed in 20 min (%)` = pct_consumed)
knitr::kable(shown, digits = c(0, 1, 2), row.names = FALSE,
             caption = "Fraction of substrate consumed during the incubation (HLM).")
cat(sprintf("Worst case over the studied range: %.2f%% consumed, at %.0f umol/L (%s)\n",
            worst$pct_consumed, worst$S, worst$system))

stopifnot(
  # Deterministic model, so this is a hard bound, not a quantile.
  max(depletion$pct_consumed) < 5,
  # Equivalently: max(v/[S]) < 4, which is exactly why Figure 2's x axis
  # runs to 4. The form as typeset would put v/[S] near 20 and could not
  # have been plotted on that axis at all.
  max(depletion$Cc / depletion$S) < 4
)
```

That last equivalence is a fourth check on the reading of Equation 2. Both
panels of source Figure 2 use an Eadie-Hofstee axis running from 0 to 4, with
the HLM points reaching roughly 3.8 and the rUGT1A4 points roughly 2.9.

```{r eh-axis-gate}
eh_range <- bind_rows(
  data.frame(system = c("HLM101", "rUGT1A4"), reading = "as typeset",
             `max v/[S]` = c(max(v_printed(Sobs2 <- seq(10, 800, by = 0.25), 445, 46, 58, 18.3, 2.3) / Sobs2),
                             max(v_printed(Sobs2, 427, 64, 79, 14.9, 2.6) / Sobs2)),
             check.names = FALSE),
  data.frame(system = c("HLM101", "rUGT1A4"), reading = "with Ks^n, Ki^n",
             `max v/[S]` = c(max(v_hill(Sobs2, 445, 46, 58, 18.3, 2.3) / Sobs2),
                             max(v_hill(Sobs2, 427, 64, 79, 14.9, 2.6) / Sobs2)),
             check.names = FALSE)
)
eh_range$`fits the 0-4 axis of Figure 2` <- eh_range$`max v/[S]` < 4
knitr::kable(eh_range, digits = 2,
             caption = "Maximum Eadie-Hofstee ordinate under each reading of Equation 2.")
stopifnot(
  all(eh_range$`max v/[S]`[eh_range$reading == "with Ks^n, Ki^n"] < 4),
  all(eh_range$`max v/[S]`[eh_range$reading == "as typeset"] > 15)
)
```

## The high-substrate asymptote

As substrate saturates both sites, every term but the doubly-occupied one drops
out and the rate tends to `Vmax * beta`. That is the mechanistic meaning of the
fixed `beta = 0.01`: the doubly-occupied enzyme still turns over, at 1% of the
fully-saturated rate. It is why the published curves decay toward a small
non-zero value rather than to zero.

```{r asymptote}
asym <- data.frame(
  system = c("HLM101", "rUGT1A4"),
  `Vmax x beta` = c(445 * 0.01, 427 * 0.01),
  `v at 100000 umol/L` = c(v_hill(1e5, 445, 46, 58, 18.3, 2.3),
                           v_hill(1e5, 427, 64, 79, 14.9, 2.6)),
  check.names = FALSE
)
knitr::kable(asym, digits = 3, caption = "The rate law tends to Vmax * beta at saturating substrate.")
stopifnot(all(abs(asym[["v at 100000 umol/L"]] / asym[["Vmax x beta"]] - 1) < 0.02))
```

# In vivo urinary MDZG

The in vivo arm of the paper is descriptive mass-balance accounting rather than
a fitted pharmacokinetic model, so it is not encoded in either model file. It is
reproduced here because it is the paper's headline result -- the first
demonstration that this pathway operates in people.

```{r urine}
urine <- tibble::tribble(
  ~Subject, ~`3 mg oral (ug)`, ~`1 mg i.v. (ug)`,
  "3",  56.3, 40.7,
  "4",  21.0, 20.8,
  "5", 148.2, 59.8,
  "6",  59.2, 17.6,
  "10", 37.1, 26.5,
  "11", 35.3, 35.1,
  "Geometric mean", 48.8, 26.5,
  "SEM", 13.2, 5.7
)
knitr::kable(urine, caption = "Source Table 3: MDZG recovered in urine over 24 h in six healthy men (reproduced verbatim, including the i.v. summary row -- see below).")
```

**The i.v. geometric mean printed in Table 3 is wrong.** Recomputing it from the
six tabulated subject values gives 30.6 ug, not 26.5 -- and 30.6 is what the
Results text ("an average of 30.7 +/- 5.7 ug") and the Discussion ("49 ug and
31 ug") both report. The printed 26.5 is subject 10's own i.v. value, so it looks
like a copy-down slip in the table. The oral column is self-consistent.

```{r urine-check}
subj <- urine[1:6, ]
gm <- function(x) exp(mean(log(x)))
knitr::kable(
  data.frame(
    Route = c("3 mg oral", "1 mg i.v."),
    `Table 3 prints` = c(48.8, 26.5),
    `Recomputed geometric mean` = round(c(gm(subj$`3 mg oral (ug)`), gm(subj$`1 mg i.v. (ug)`)), 1),
    `Results text` = c("48.8", "30.7"),
    check.names = FALSE
  ),
  row.names = FALSE,
  caption = "The oral summary reproduces; the i.v. one does not."
)
stopifnot(
  abs(gm(subj$`3 mg oral (ug)`) - 48.8) < 0.2,   # oral row is self-consistent
  abs(gm(subj$`1 mg i.v. (ug)`) - 30.7) < 0.2,   # i.v. matches the TEXT, not the table
  abs(gm(subj$`1 mg i.v. (ug)`) - 26.5) > 3      # ... and is genuinely far from the printed value
)
```

Nothing in either model file depends on this row; it is recorded because the
source trace has to be honest about which printed numbers reproduce.

MDZG accounted for roughly 1-2% of the administered dose on a molar basis, with
more than 70% of it excreted in the first 12 h under both routes. Between-subject
variability was about sevenfold after oral dosing and 3.4-fold after intravenous
dosing -- the same sevenfold spread the authors saw across a 56-donor bank of
liver microsomes.

# Assumptions and deviations

* **Equation 2 is implemented with $K_s^n$ and $K_i^n$, not as typeset.** The
  form printed in the paper is dimensionally inconsistent, reproduces neither
  Figure 3 nor the Figure 4 control rates, and disagrees with the Shou *et al.*
  model the Methods cite as its source. All three checks are run above as
  render-time gates. This is the single deviation from the printed source, and
  it is a correction of the typesetting, not of the model. Both model files
  carry the same reasoning inline.
* **Known deviation: the two curves cross between about 150 and 345 umol/L.**
  Source Figure 3 keeps the HLM curve above the rUGT1A4 curve across the whole
  descending limb. The implemented pair cross twice, with rUGT1A4 up to 3.2%
  (about 5 pmol/min/mg) above HLM in that window; outside it the published
  ordering is reproduced, reaching a 20% gap by 800 umol/L. Table 4 reports
  `alpha` (18.3 vs 14.9) and `n` (2.3 vs 2.6) to two or three significant
  figures, and those two parameters dominate the curve shape, so a sub-5%
  vertical offset between two nearly-coincident curves is within the published
  precision. No parameter was adjusted to remove it; the ordering gate is
  asserted only in the tail, where the published separation is large enough to
  be meaningful.
* **The rate law is written multiplied through by $[S]^n$.** Algebraically
  identical to the printed ratio, but finite at $[S] = 0$, where the printed
  form evaluates $1/0$. This is also how Shou *et al.* write the model.
* **No residual error.** The paper reports fit quality only as $R^2$ (0.962 for
  HLM, 0.986 for rUGT1A4) and gives no residual standard deviation, so `addSd`
  is `fixed(0)` rather than invented. Both models are deterministic: they carry
  no random effects at all, which is why the assertions above are tight rather
  than quantile-based.
* **No inter-individual variability.** The HLM preparation is a pool and the
  recombinant enzyme is a single expressed protein; neither fit has a subject
  dimension. The paper separately notes a sevenfold range in MDZG formation
  across a 56-donor microsome bank, but reports no variance component for it.
* **The ketoconazole, itraconazole and trifluoperazine effects are not
  encoded.** Source Figure 4 reports IC50 values of 150 +/- 6 umol/L
  (ketoconazole) and 308 +/- 14 umol/L (itraconazole), and a maximal
  trifluoperazine activation of 138% at 60 umol/L. No inhibition equation, Hill
  slope or mechanism is published for any of the three. The fitted sigmoids in
  Figure 4 are visibly steeper than a slope-1 curve, so encoding them with an
  assumed slope of 1 would misrepresent the published data, and inventing a
  slope is not permitted. The values are recorded in each model file's
  `population$notes` instead. The authors' own conclusion is that this
  inhibition is weak enough to be irrelevant in vivo.
* **The Simcyp simulation is out of scope.** The Discussion reports a
  population-ADME simulation (Simcyp v8.1, 10 populations of 10 individuals,
  3 mg oral midazolam) in which UGT1A4 contributes 2.9% of hepatic intrinsic
  clearance uninhibited and 19% with 400 mg ketoconazole co-administered. That
  simulation runs on "the default MDZ profile provided by Simcyp", a
  proprietary platform parameter set that is not in the paper or any supplement.
  Platform PBPK models are out of scope for this library and none of those
  parameters may be substituted from elsewhere, so only the enzyme-kinetic layer
  the paper itself reports is extracted.
* **PKNCA is not used.** There is no dose, no absorption-distribution-
  elimination profile and no concentration-time curve to integrate; this is an
  enzyme-kinetic model, so the validation follows the mechanistic pattern
  (dimensional analysis, mass balance, limiting behaviour, figure replication)
  rather than non-compartmental analysis.
* **"Km" versus `Ks`.** The Abstract and Discussion call the productive-binding
  dissociation constant `Km` (46 and 64 umol/L); Table 4 and Equation 2 call it
  `Ks`. They are the same quantity. The model files use `lks`, matching the
  equation.
* **No parameter was digitised.** Every value in both `ini()` blocks comes from
  Table 4 or from the Results text. Figures 3 and 4 are used only to *check* the
  implementation, never to source a value.
