---
title: "Nicorandil (Iida 2008)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nicorandil (Iida 2008)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Iida S, Kinoshita H, Holford NHG. Population pharmacokinetic and
  pharmacodynamic modelling of the effects of nicorandil in the treatment of
  acute heart failure. Br J Clin Pharmacol 66(3):352-365, 2008.
  doi:10.1111/j.1365-2125.2008.03257.x.
* Indication: acute heart failure (AHF), including acute exacerbation of
  chronic heart failure. NYHA class II-IV. Nicorandil IV is approved for AHF
  in Japan (oral nicorandil is also used for unstable angina across Europe
  and Asia).
* Article: <https://doi.org/10.1111/j.1365-2125.2008.03257.x>

# Population

Five clinical studies were pooled: two studies in 11 healthy adult Japanese
male volunteers (concentration-only data, no haemodynamic measurements) and
three studies in 94 AHF patients in whom both plasma nicorandil concentration
and pulmonary artery wedge pressure (PAWP) were measured via a Swan-Ganz
catheter (Iida 2008 Table 1). 618 nicorandil concentrations and 559 PAWP
observations contributed to the joint model. The AHF cohort was older (study
means 62.8-70.8 years vs 33.0-41.2 years in healthy volunteers) and
lighter (study means 58.6-61.8 kg vs 62.6-63.6 kg in healthy). AHF
aetiologies pooled across studies were ischaemic heart disease (33), valvular
heart disease (22), dilated cardiomyopathy (22), and hypertensive heart
disease (13). Inclusion required baseline PAWP greater than 18 mmHg in the
bolus and bolus+infusion AHF studies and greater than 15 mmHg in the
long-term infusion study. Baseline PAWP at predose (mean +/- SD) was 25.9
+/- 6.4 mmHg, 26.9 +/- 6.3 mmHg, and 27.2 +/- 7.4 mmHg in the three AHF
studies respectively (Table 1).

The same population metadata is available programmatically via
`readModelDb("Iida_2008_nicorandil")$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Iida_2008_nicorandil.R`. The table below
collects the trace in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` = log(26.3 * 1.94) | CL = 26.3 L/h/70kg (healthy); FCL = 1.94 (AHF) | Table 2 POP_CL row + FCL row |
| `lvc` = log(18.1 * 1.39) | V1 = 18.1 L/70kg (healthy); FV1 = 1.39 (AHF) | Table 2 POP_V1 row + FV1 row |
| `lq`  = log(71.6 * 0.519) | Q  = 71.6 L/h/70kg (healthy); FQ  = 0.519 (AHF) | Table 2 POP_Q row + FQ row |
| `lvp` = log(24.1 * 4.06) | V2 = 24.1 L/70kg (healthy); FV2 = 4.06 (AHF) | Table 2 POP_V2 row + FV2 row |
| `e_dis_healthy_cl` = log(1 / 1.94) | -0.6627: shifts AHF typical to healthy | Table 2 FCL = 1.94 |
| `e_dis_healthy_vc` = log(1 / 1.39) | -0.3293: shifts AHF typical to healthy | Table 2 FV1 = 1.39 |
| `e_dis_healthy_q`  = log(1 / 0.519)| +0.6553: shifts AHF typical to healthy | Table 2 FQ  = 0.519 |
| `e_dis_healthy_vp` = log(1 / 4.06) | -1.4015: shifts AHF typical to healthy | Table 2 FV2 = 4.06 |
| `e_wt_cl` = fixed(0.75) | Allometric exponent on CL | Equation 5 (paper p. 356): exponent 3/4 |
| `e_wt_vc` = fixed(1.0) | Allometric exponent on V1 | Holford 1996 [11]; canonical 1.0 on volumes |
| `e_wt_q`  = fixed(0.75) | Allometric exponent on Q | Equation 5 covers all clearance terms |
| `e_wt_vp` = fixed(1.0) | Allometric exponent on V2 | Holford 1996 [11]; canonical 1.0 on volumes |
| `lrbase` = log(25.6) | S0 = 25.6 mmHg | Table 2 POP_S0 row |
| `lsss` = log(19.5) | Sss = 19.5 mmHg | Table 2 POP_Sss row |
| `ltprog` = log(5.83) | Tprog = 5.83 h | Table 2 POP_TPROG row |
| `limax` = log(11.7) | Magnitude of inhibitory Emax | Table 2 POP_Emax = -11.7 mmHg |
| `lec50` = log(423) | EC50 = 423 ng/mL | Table 2 POP_EC50 = 423 ug/L |
| PK IIV block etalcl + etalvc + etalq | variances 0.686 / 0.301 / 0.414; corr R78 = 0.257, R89 = 0.957, R79 fixed 0 | Table 4 BSV_CL / BSV_V1 / BSV_Q + R78 / R89 |
| PD IIV block etalrbase + etalsss + etalec50 | variances 0.210 / 0.320 / 0.879; corr R12 = 0.864, R23 = 0.054, R13 fixed 0 | Table 4 PPV_S0 / PPV_Sss / PPV_EC50 + R12 / R23 |
| `propSd` = 0.212 | Proportional residual error 21.2% | Table 4 RUV_CVCP row |
| `addSd`  = 6.82  | Additive residual error 6.82 ng/mL | Table 4 RUV_SDCP row |
| `addSd_pawp` = 2.5 | Additive residual error 2.5 mmHg on PAWP | Table 4 RUV_SDFX row |
| `d/dt(central) ...` two-compartment IV PK | Two-compartment disposition | Results paragraph 'Pharmacokinetics of nicorandil' |
| `progress` = (Sss - S0) * (1 - exp(-log(2)/Tprog * t)) | Asymptotic-exponential disease progression | Equation 4 (paper p. 355) |
| `pdeff` = -Imax * Cc / (EC50 + Cc) | Inhibitory Emax drug effect on PAWP | Equation 2 with Emax negative |
| `pawp` = S0 + progress + pdeff | Final PD observation equation | Equation 4 with PD(t) = pdeff |

The DIS_HEALTHY re-expression of the AHF-vs-healthy contrast is detailed
inline in the `covariateData[["DIS_HEALTHY"]]$notes` entry of the model file
and in the corresponding entry of `inst/references/covariate-columns.md`.

# Virtual cohort

Original observed data are not publicly available. The simulation below uses
a small virtual cohort whose covariate distribution approximates the AHF
patient cohort in the Iida 2008 trial: 60 kg typical body weight per the
mean of the three AHF studies (Table 1), pooled across the bolus and
bolus+infusion regimens. The recommended dosing regimen for AHF from the
Discussion section is a 200 ug/kg IV bolus over 5 minutes followed by a 400
ug/kg/h IV infusion; we simulate this regimen alongside an unrelated 12 mg IV
bolus (for comparison with paper Figure 2 left upper panel).

```{r cohort}
set.seed(20260613L)

n_ahf <- 200L
typical_wt <- 60   # kg, AHF cohort mean (Table 1)

make_ahf_cohort <- function(n, regimen, id_offset = 0L) {
  weights <- pmin(pmax(rlnorm(n, meanlog = log(typical_wt), sdlog = 0.15), 40), 95)
  tibble::tibble(
    id           = id_offset + seq_len(n),
    WT           = weights,
    DIS_HEALTHY  = 0L,   # AHF reference cohort
    regimen      = regimen
  )
}

cohort_bolus <- make_ahf_cohort(n_ahf, regimen = "12 mg bolus",                 id_offset =       0L)
cohort_reg   <- make_ahf_cohort(n_ahf, regimen = "200 ug/kg + 200 ug/kg/h 24 h", id_offset =   n_ahf)
cohort_rec   <- make_ahf_cohort(n_ahf, regimen = "200 ug/kg + 400 ug/kg/h 24 h", id_offset = 2*n_ahf)

build_events <- function(cohort_df, regimen) {
  if (regimen == "12 mg bolus") {
    dose_times <- 0
    dose_amts  <- rep(12, nrow(cohort_df))            # mg, fixed dose per subject
    dose_rates <- rep(NA_real_, nrow(cohort_df))
    end_obs    <- 6                                    # h
  } else if (regimen == "200 ug/kg + 200 ug/kg/h 24 h") {
    bolus_mg   <- 200 / 1000 * cohort_df$WT            # 200 ug/kg -> mg
    infus_mg_h <- 200 / 1000 * cohort_df$WT            # 200 ug/kg/h -> mg/h
    dose_times <- c(0)
    dose_amts  <- bolus_mg
    dose_rates <- infus_mg_h
    end_obs    <- 30
  } else {
    bolus_mg   <- 200 / 1000 * cohort_df$WT            # 200 ug/kg -> mg
    infus_mg_h <- 400 / 1000 * cohort_df$WT            # 400 ug/kg/h -> mg/h
    dose_times <- c(0)
    dose_amts  <- bolus_mg
    dose_rates <- infus_mg_h
    end_obs    <- 30
  }
  obs_t <- sort(unique(c(seq(0, end_obs, by = 0.25), seq(end_obs, 48, by = 1))))
  do.call(rbind, lapply(seq_len(nrow(cohort_df)), function(i) {
    s     <- cohort_df[i, ]
    rows  <- list()
    rows[[1]] <- data.frame(
      id = s$id, time = 0, amt = if (regimen == "12 mg bolus") 12 else (200/1000)*s$WT,
      evid = 1L, cmt = "central", rate = NA_real_,
      WT = s$WT, DIS_HEALTHY = s$DIS_HEALTHY, regimen = s$regimen,
      stringsAsFactors = FALSE
    )
    if (regimen != "12 mg bolus") {
      # 24 h constant infusion at the regimen's mg/h rate, started at t = 0.
      rows[[2]] <- data.frame(
        id = s$id, time = 0,
        amt  = (if (regimen == "200 ug/kg + 200 ug/kg/h 24 h") 200 else 400) / 1000 * s$WT * 24,
        evid = 1L, cmt = "central",
        rate = (if (regimen == "200 ug/kg + 200 ug/kg/h 24 h") 200 else 400) / 1000 * s$WT,
        WT = s$WT, DIS_HEALTHY = s$DIS_HEALTHY, regimen = s$regimen,
        stringsAsFactors = FALSE
      )
    }
    rows[[length(rows) + 1]] <- data.frame(
      id = s$id, time = obs_t, amt = 0, evid = 0L, cmt = "Cc", rate = 0,
      WT = s$WT, DIS_HEALTHY = s$DIS_HEALTHY, regimen = s$regimen,
      stringsAsFactors = FALSE
    )
    rows[[length(rows) + 1]] <- data.frame(
      id = s$id, time = obs_t, amt = 0, evid = 0L, cmt = "pawp", rate = 0,
      WT = s$WT, DIS_HEALTHY = s$DIS_HEALTHY, regimen = s$regimen,
      stringsAsFactors = FALSE
    )
    do.call(rbind, rows)
  }))
}

events <- rbind(
  build_events(cohort_bolus, "12 mg bolus"),
  build_events(cohort_reg,   "200 ug/kg + 200 ug/kg/h 24 h"),
  build_events(cohort_rec,   "200 ug/kg + 400 ug/kg/h 24 h")
)
events <- events[order(events$id, events$time, -events$evid), ]
rownames(events) <- NULL
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid", "cmt")])))
```

# Simulation

```{r simulate}
mod <- readModelDb("Iida_2008_nicorandil")
sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep   = c("regimen", "WT", "DIS_HEALTHY")
) |>
  as.data.frame()
```

Typical-value (between-subject-variability-zeroed) replication for the
figure comparisons below:

```{r simulate-typical}
mod_typical <- mod |> rxode2::zeroRe()
sim_typical <- rxode2::rxSolve(
  mod_typical,
  events = events,
  keep   = c("regimen", "WT", "DIS_HEALTHY")
) |>
  as.data.frame()
```

# Replicate published figures

## Figure 2 -- Concentration profiles in AHF patients (bolus regimen)

Iida 2008 Figure 2 left upper panel shows individual plasma nicorandil
concentration vs time after IV bolus doses of 4, 8, 12, and 18 mg in AHF
patients. Concentrations decline rapidly within 1-2 h. Below we replicate
the 12 mg bolus arm as a population-VPC band (median and 5th-95th
percentile of the 200-subject virtual cohort) plus the typical-value
trajectory.

```{r figure-2, fig.width = 7, fig.height = 4}
sim_bolus_vpc <- sim |>
  dplyr::filter(regimen == "12 mg bolus", !is.na(Cc), time <= 6) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

sim_bolus_typ <- sim_typical |>
  dplyr::filter(regimen == "12 mg bolus", !is.na(Cc), time <= 6, id == min(id))

ggplot(sim_bolus_vpc, aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.20, fill = "steelblue") +
  geom_line(colour = "steelblue") +
  geom_line(data = sim_bolus_typ, aes(time, Cc), colour = "black", linetype = "dashed") +
  scale_y_log10() +
  labs(
    x = "Time after bolus (h)",
    y = "Nicorandil concentration (ng/mL, log scale)",
    title = "AHF 12 mg IV bolus",
    caption = "VPC band (5th-95th percentile, blue ribbon) and typical-value trajectory (dashed); replicates the 12 mg arm of Figure 2 left upper panel of Iida 2008."
  )
```

## Figure 7 -- PAWP after the recommended AHF regimen

Iida 2008 Discussion recommends a 200 ug/kg loading dose followed by a 400
ug/kg/h continuous infusion to achieve approximately a 30% reduction in
PAWP. Below we plot the simulated PAWP trajectory for the typical AHF
patient under that regimen and a comparator at 200 ug/kg/h infusion. The
expected baseline is S0 = 25.6 mmHg and the asymptotic disease-progression-
only value is Sss = 19.5 mmHg, so a 30% reduction relative to S0 lands at
approximately 17.9 mmHg.

```{r figure-7, fig.width = 7, fig.height = 4}
sim_pd_typ <- sim_typical |>
  dplyr::filter(regimen != "12 mg bolus", !is.na(pawp), time <= 24) |>
  dplyr::group_by(regimen) |>
  dplyr::filter(id == min(id)) |>
  dplyr::ungroup()

ggplot(sim_pd_typ, aes(time, pawp, colour = regimen, linetype = regimen)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 25.6 * 0.7, linetype = "dotted", colour = "grey50") +
  annotate("text", x = 0.5, y = 25.6 * 0.7 + 0.4,
           label = "30% reduction target (~17.9 mmHg)",
           hjust = 0, size = 3, colour = "grey40") +
  coord_cartesian(ylim = c(10, 30)) +
  labs(
    x = "Time (h)",
    y = "PAWP (mmHg)",
    title = "Typical AHF patient: PAWP trajectory by infusion rate",
    caption = "Replicates the haemodynamic concept of Figure 7 of Iida 2008 / Discussion: the recommended 200 ug/kg + 400 ug/kg/h regimen achieves ~30% PAWP reduction."
  )
```

# PKNCA validation

Iida 2008 does not report a formal NCA table; the paper's quantitative PK
inferences derive from the popPK structural parameters in Table 2 rather
than from a separate NCA summary. We therefore use PKNCA here as a model
self-check: extract Cmax and AUC from the simulated nicorandil
concentration profiles after the 12 mg AHF bolus and verify the magnitudes
against the closed-form predictions implied by the model.

```{r pknca}
# The event table requests two observation compartments per subject (Cc and
# pawp); rxSolve emits both as separate rows with identical (id, time, Cc)
# values. Deduplicate so PKNCA sees one concentration row per (id, time).
sim_nca <- sim |>
  dplyr::filter(regimen == "12 mg bolus", !is.na(Cc), time <= 6) |>
  dplyr::select(id, time, Cc, regimen) |>
  dplyr::distinct(id, time, regimen, .keep_all = TRUE)

# Time-zero guarantee per the pknca-recipes 'Time-zero records (mandatory)'
# note. For an IV bolus, the t = 0 row already comes back with Cc set to the
# initial Dose / V1, so the bind_rows() pattern of the template would
# overwrite a non-zero initial concentration with 0; instead we just verify
# every (id, time = 0) row is already present after deduplication.
sim_nca <- sim_nca |>
  dplyr::group_by(id) |>
  dplyr::mutate(has_t0 = any(time == 0)) |>
  dplyr::ungroup()
stopifnot(all(sim_nca$has_t0))
sim_nca$has_t0 <- NULL

conc_obj <- PKNCA::PKNCAconc(
  sim_nca,
  Cc ~ time | regimen + id,
  concu = "ng/mL",
  timeu = "h"
)

dose_df <- events |>
  dplyr::filter(regimen == "12 mg bolus", evid == 1L) |>
  dplyr::select(id, time, amt, regimen)

dose_obj <- PKNCA::PKNCAdose(
  dose_df,
  amt ~ time | regimen + id,
  doseu = "mg"
)

intervals <- data.frame(
  start      = 0,
  end        = 6,
  cmax       = TRUE,
  tmax       = TRUE,
  auclast    = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- PKNCA::pk.nca(nca_data)
```

## Comparison against closed-form predictions

For a 12 mg IV bolus given to a typical AHF patient (60 kg), the model
predicts Cmax = Dose / V1 (with V1 scaled allometrically by weight). The
typical AHF V1 at 60 kg is V1_AHF * (60/70)^1.0 = 18.1 * 1.39 * (60/70) =
21.57 L, giving Cmax = 12 mg / 21.57 L = 556 ng/mL. The simulated AUC0-6
should match the trapezoidal integral of the two-compartment IV profile.

```{r nca_compare}
predicted <- tibble::tibble(
  regimen     = "12 mg bolus",
  cmax        = 556,    # ng/mL; closed-form Dose / V1_AHF_60kg
  tmax        = 0,
  auclast     = NA_real_
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = predicted,
  by            = "regimen",
  units         = c(cmax = "ng/mL", auclast = "ng*h/mL", tmax = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated vs closed-form predicted NCA. * differs from prediction by >20%.",
  align   = c("l", "l", "r", "r", "r")
)
```

The simulated VPC-median Cmax should sit within roughly +-20% of the
closed-form prediction. The closed-form does not predict tmax (a bolus
peaks at t = 0) or AUC (no closed form for the 6-h truncated trapezoid),
so those cells are NA in the reference.

# Assumptions and deviations

- **AHF-vs-healthy disease cohort encoded via the canonical DIS_HEALTHY
  covariate (reverse-coded).** Iida 2008 Table 2 reports POP_CL, POP_V1,
  POP_Q, POP_V2 as healthy-volunteer typical values with multiplicative
  AHF/healthy factors FCL = 1.94, FV1 = 1.39, FQ = 0.519, FV2 = 4.06. The
  canonical `DIS_HEALTHY` orientation in nlmixr2lib has the patient cohort
  as the reference (DIS_HEALTHY = 0) and the healthy cohort as the +shift
  (DIS_HEALTHY = 1). The model file therefore re-anchors the typical-value
  parameters on the AHF reference state by multiplying the published
  healthy typicals by the corresponding F factor, and sets the
  `e_dis_healthy_<pk>` covariate-effect parameters to `log(1 / F_X)` so
  DIS_HEALTHY = 1 recovers the published healthy-volunteer typical. This
  re-expression follows the Bulitta_2010_ceftazidime.R precedent and is
  documented on the `DIS_HEALTHY` entry of
  `inst/references/covariate-columns.md`. PK predictions for individual
  patients are unaffected; only the structural-typical reading of the
  `ini()` parameters shifts from the healthy-cohort interpretation in the
  paper's Table 2 to the AHF-cohort interpretation in the model file.
- **Allometric exponents fixed at 0.75 on CL and Q, 1.0 on V1 and V2.**
  Iida 2008 Equation 5 writes the allometric scaling explicitly with
  exponent 3/4 for CL ("CL_GRP = CL_POPSTD * (Wt / 70)^(3/4)") and the
  prose says "Differences in CL, Q, V1 and V2 were assumed to be explained
  in part by an allometric function of weight [11] e.g. Equation 5,"
  citing Holford 1996 (Clin Pharmacokinet 30:329-32) as reference [11].
  We therefore use 0.75 for CL and Q (clearance terms) and 1.0 for V1 and
  V2 (volumes) per the Holford size-standard convention; the paper does
  not separately tabulate the volume exponent, but the per-70-kg units in
  Table 2 (l 70 kg^-1) match this convention.
- **Reference weight 70 kg.** Stated explicitly in Methods paragraph
  'Covariate effects': "Size differences were based on a standard weight of
  70 kg." The cohort mean weight across the three AHF studies was
  approximately 60 kg.
- **No effect compartment.** Iida 2008 tested an effect-compartment
  parameterization for the PD layer and rejected it: the estimated
  effect-compartment half-life was 2.2 min with very small variability;
  the NONMEM run finished with rounding errors and the objective function
  improved by only 8.9. The final model 10 used the plasma concentration
  directly as the driver of the immediate-effect Emax model. We follow the
  paper's choice and do not include an effect compartment.
- **Inhibitory Emax sign convention.** The paper reports POP_Emax = -11.7
  mmHg (negative, representing PAWP reduction). The model file stores the
  magnitude as `limax = log(11.7)` (a positive log-magnitude in the `ini()`
  block) and the negative sign is applied explicitly in `model()` via
  `pdeff = -imax * Cc / (ec50 + Cc)`. This keeps `limax` on a stable
  log-transformed scale for any downstream re-estimation.
- **CL-Q covariance (R79) fixed to 0; S0-EC50 covariance (R13) fixed to
  0.** Iida 2008 Table 4 reports R78 (CL-V1, 0.257) and R89 (V1-Q, 0.957)
  on the PK side and R12 (S0-Sss, 0.864) and R23 (Sss-EC50, 0.054) on the
  PD side. The third pairwise correlations (R79 and R13) are not reported
  in the table. We interpret the absence as having been fixed to zero in
  the OMEGA block of the final NONMEM run (the alternative interpretation
  is that they were estimated and found to be essentially zero) and set
  the corresponding off-diagonal entries to `fixed(0)` in the model file's
  IIV block.
- **Between-occasion variability (BOV) on CL omitted.** Iida 2008 Table 4
  reports BOV_CL1 = 0.199 (between-occasion CL variability variance,
  paper Equation 7 with occasion-indexed kappa). nlmixr2 has no native
  mechanism for arbitrary-occasion BOV without enumerating per-occasion
  etas across the longest expected dosing history; we therefore omit the
  BOV term in this implementation. Population-level typical-value
  predictions (the trajectories plotted above) are unaffected; the
  between-subject random-effect spread in the VPC band underestimates the
  per-subject within-subject variability across the multi-occasion
  studies by the amount BOV would contribute.
- **eta-on-residual-error scaling omitted.** Iida 2008 Equation 8
  introduces a per-subject random scaling `exp(eta_PPV_RUV)` on the
  residual error magnitude (paper Table 4 PPV_RUVCP = 0.716 for
  nicorandil and PPV_RUVFX = 0.308 for PAWP, citing Karlsson, Jonsson,
  Wiltse, Wade 1998 [14]). nlmixr2 does not have a native parameterization
  for eta-on-sigma residual scaling; we use the typical residual error
  magnitudes (RUV_CVCP = 0.212, RUV_SDCP = 6.82 ng/mL, RUV_SDFX = 2.5
  mmHg from Table 4) without subject-level scaling. This affects only the
  estimation reweighting and the absolute spread of the simulated
  observations around the predictions; population-level trajectories are
  identical.
- **PK structural reference is healthy volunteers.** The PD layer (S0,
  Sss, Tprog, Emax, EC50) was estimated only from AHF patients (no PAWP
  measurements were collected in healthy volunteers), so DIS_HEALTHY = 1
  predictions of PAWP are not physiologically meaningful and should not
  be used. The model still computes a `pawp` output for all subjects to
  preserve a single-model interface; downstream users should sample
  `pawp` only when DIS_HEALTHY = 0.
- **Virtual cohort body weight.** We sample weight log-normally around 60
  kg (sdlog = 0.15) bracketing the per-study AHF means of 58.6, 60.3, and
  61.8 kg (Table 1). Individual-subject weights from the trials are not
  publicly available; the simulated cohort therefore matches the
  population-level summary but not the individual-level distribution.
- **DIS_HEALTHY covariate for healthy volunteer cohort prevalence is
  fixed at 0.** All simulated subjects in the cohort represent AHF
  patients (DIS_HEALTHY = 0); the model is designed to predict the
  haemodynamic response of nicorandil in the AHF cohort, which is the
  population for which the recommended dosing regimen was developed.
