---
title: "Valemetostat population PK in relapsed/refractory PTCL (Inoue 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Valemetostat population PK in relapsed/refractory PTCL (Inoue 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5, error = FALSE
)
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
```

# Model and source

Inoue 2025 reports an updated population pharmacokinetic (PPK) model for
**valemetostat**, an oral EZH1/EZH2 dual inhibitor approved in Japan for
relapsed/refractory (R/R) adult T-cell leukemia/lymphoma (ATLL) and R/R
peripheral T-cell lymphoma (PTCL), given 200 mg once daily.

* Article: <https://doi.org/10.1002/jcph.70100> (PMC12649291)
* J Clin Pharmacol. 2025;65(12):1699-1711.

The model is fitted to **4635 total** and **3085 unbound** valemetostat
concentrations from 342 participants pooled across six trials, and it describes
both analytes *simultaneously*. Three features make it structurally unusual.

**The central compartment is a binding equilibrium, not a well-stirred pool.**
Valemetostat binds alpha-1-acid glycoprotein (AAG). The state variable
`central` holds *total* drug; the unbound concentration is recovered
analytically from a single-site saturable binding sub-model with capacity
`bmax` (the paper's RMAX) and dissociation constant `kd`. Because
`Ctot = Cu + bmax * Cu / (kd + Cu)`, the unbound concentration is the positive
root of a quadratic.

**Everything downstream is driven by unbound drug.** Clearance and *both*
inter-compartmental clearances act on `Cu`, exactly as drawn in the paper's
Figure 1. This is what makes the model nonlinear in total drug even though
unbound elimination is linear.

**AAG enters three places using only two estimated exponents.** It scales the
binding capacity (exponent 0.805) and it scales CL *and* F1 through one shared
exponent (0.336). Since unbound exposure is `dose * F1 / CL`, the shared
exponent cancels: unbound exposure is algebraically independent of AAG while
total exposure still rises with it. That cancellation is the paper's central
pharmacological result and it is checked numerically below.

```{r load-model}
mod <- nlmixr2lib::modellib("Inoue_2025_valemetostat")
mod
```

# Population

```{r population}
pop <- nlmixr2lib::readModelDb("Inoue_2025_valemetostat")
```

The PPK analysis set (Inoue 2025 Table 1) is 342 participants: mean (SD) age
59.8 (17.2) years, weight 69.3 (15.9) kg, 64.3% male, and AAG 113 (62.3) mg/dL.
Disease type is PTCL 55.6%, ATLL 17.8%, other non-Hodgkin lymphoma 5.6%, with
21.1% non-patients (healthy Japanese participants from the J107 drug-drug
interaction and J109 food-effect studies, pooled with non-cancer participants
with hepatic impairment from U106). Race/country enrolled is Asian Japanese
41.8%, Asian non-Japanese 6.1%, White 39.2%, Other 12.9%. Hepatic function by
NCI-ODWG criteria is normal 74.9%, mild 21.1%, moderate 3.5%; creatinine
clearance is 89.2 (35.5) mL/min.

The covariate **reference individual** used throughout the paper's inference
(Figure 2 caption) is a male PTCL patient, Asian Japanese, on no interacting
comedication, with normal hepatic function, weighing 68.2 kg, aged 65 years,
with AAG 100 mg/dL and creatinine clearance 83 mL/min. Every reference value in
the model file is taken from that sentence.

```{r reference-individual}
refCov <- data.frame(
  WT = 68.2, AGE = 65, CRCL = 83, AAG = 100,
  SEXF = 0, TUMTP_ATLL = 0, DIS_BCELLNHL = 0, DIS_HEALTHY = 0,
  RACE_ASIAN_OTH = 0, RACE_WHITE = 0, RACE_OTHER = 0,
  HEPIMP_MILD = 0, HEPIMP_MOD = 0,
  CONMED_PGP_INH = 0, CONMED_CYP3A4_INH = 0, STUDY_J101 = 0
)
knitr::kable(t(refCov), col.names = "Reference individual",
             caption = "Covariate reference values (Inoue 2025 Figure 2 caption).")
```

# A note on units

Inoue 2025 Table 2 prints `KD` and `RMAX` in **nmol/L**, and the paper nowhere
states a molar mass for valemetostat. The binding sub-model therefore can only
be written on a molar scale without importing a constant from outside the
source, so this model file declares `dosing = "nmol"` and
`concentration = "nmol/L"`, and doses below are in nmol.

That choice costs nothing in validation power, because every check in this
vignette is either a **ratio** (unit-free) or the **mass-balance identity**
`Cu,ss,avg = dose rate / CL`, which holds in whatever mass unit the dose is
supplied in. In particular, applying that identity in mg gives

```{r mg-arithmetic}
c(`200 mg / 24 h / 520 L/h, in ng/mL` = 200 / 24 / 520 * 1000)
```

which brackets the paper's own reference unbound average concentrations of
**13.9 ng/mL** (efficacy analysis, Figure S6 caption) and **18.1 ng/mL**
(safety analysis, Figure S8 caption). Mapping a milligram dose onto the molar
scale of the binding sub-model would need the molar mass; see *Assumptions and
deviations*.

# Source trace

```{r source-trace}
sourceTrace <- tibble::tribble(
  ~Quantity,                          ~Value,     ~`Source location`,
  "CL (unbound), L/h",                "520",      "Table 2, 'CL/F, L/h'",
  "V1, L",                            "42.5",     "Table 2, 'V1/F, L'",
  "Q2, L/h",                          "40.0",     "Table 2, 'Q2/F, L/h'",
  "V2, L",                            "3670",     "Table 2, 'V2/F, L'",
  "Q3, L/h",                          "230",      "Table 2, 'Q3/F, L/h'",
  "V3, L",                            "2950",     "Table 2, 'V3/F, L'",
  "KA, 1/h",                          "0.374",    "Table 2, 'KA, 1/h'",
  "D1 (zero-order duration), h",      "1.22",     "Table 2, 'D1, h'",
  "KD, nmol/L",                       "221",      "Table 2, 'KD, nmol/L'",
  "RMAX (bmax), nmol/L",              "8280",     "Table 2, 'RMAX, nmol/L'",
  "J101 unbound assay factor",        "0.638",    "Table 2, 'ASSAY'",
  "F1",                               "1 (FIXED)","Table 2, 'F1, Relative bioavailability'",
  "Allometric exponent, clearances",  "0.75 (FIXED)", "Table 2, 'CL ~ WT'",
  "Allometric exponent, volumes",     "1 (FIXED)",    "Table 2, 'V ~ WT'",
  "AAG exponent on bmax",             "0.805",    "Table 2, 'RMAX ~ AAG'",
  "AAG exponent on CL and on F1",     "0.336",    "Table 2, 'CL/F ~ AAG' and 'F1 ~ AAG' (one common parameter)",
  "Age exponent on CL",               "-0.205",   "Table 2, 'CL/F ~ AGE'",
  "CrCl exponent on CL",              "-0.0107",  "Table 2, 'CL/F ~ CrCl'",
  "CL ratio, ATLL",                   "0.828",    "Table 2, 'CL/F ~ ATLL'",
  "CL ratio, other NHL",              "0.847",    "Table 2, 'CL/F ~ OTHER NHL'",
  "CL ratio, non-patient",            "0.942",    "Table 2, 'CL/F ~ NON-PATIENT'",
  "CL ratio, female",                 "1.06",     "Table 2, 'CL/F ~ FEMALE'",
  "CL ratio, Asian non-Japanese",     "1.25",     "Table 2, 'CL/F ~ ASIAN NON-J'",
  "CL ratio, White",                  "0.967",    "Table 2, 'CL/F ~ WHITE'",
  "CL ratio, other race",             "0.816",    "Table 2, 'CL/F ~ OTHER RACE'",
  "CL ratio, mild hepatic impairment","0.892",    "Table 2, 'CL/F ~ HEPAT MILD'",
  "CL ratio, moderate hepatic impairment","0.752","Table 2, 'CL/F ~ HEPAT MOD'",
  "F1 ratio, P-gp inhibitor",         "1.29",     "Table 2, 'F1 ~ DDI PGPi'",
  "F1 ratio, CYP3A inhibitor",        "1.23",     "Table 2, 'F1 ~ DDI CYPi PGPi'",
  "Omega block CL/V1",                "0.249, 0.613, 2.49", "Table 2, Omega(1,1), Omega(2,1), Omega(2,2)",
  "Omega block KA/D1",                "0.0563, 0.0629, 1.65","Table 2, Omega(3,3), Omega(4,3), Omega(4,4)",
  "Omega bmax",                       "0.0564",   "Table 2, Omega(5,5)",
  "Omega F1",                         "0.485",    "Table 2, Omega(6,6)",
  "Sigma, total",                     "0.388",    "Table 2, 'Log-additive - Total'",
  "Sigma, unbound non-J101",          "0.404",    "Table 2, 'Log-additive - Unbound/non-DS3201-A-J101'",
  "Sigma, unbound J101",              "0.327",    "Table 2, 'Log-additive - Unbound/DS3201-A-J101'",
  "Covariate reference values",       "68.2 kg, 65 y, AAG 100 mg/dL, CrCl 83 mL/min", "Figure 2 caption",
  "Model structure (schematic)",      "3-cmt, zero+first-order absorption, saturable binding", "Figure 1"
)
knitr::kable(sourceTrace, caption = "Source trace for every ini() value and structural feature.")
```

## Reporting-scale conventions, settled from the table's own derived columns

Table 2 prints two derived columns that pin down the variance conventions
without any assumption on our part. The OMEGA rows report **log-scale
variances**: the printed `CV%` reproduces as `sqrt(exp(var) - 1)` and the
printed `Corr` reproduces as `cov / sqrt(var1 * var2)`. The SIGMA rows use a
*different* convention: their printed `CV%` is simply `sqrt(var)`.

```{r scale-conventions}
omegaVar  <- c(CL = 0.249, V1 = 2.49, KA = 0.0563, D1 = 1.65, RMAX = 0.0564, F1 = 0.485)
sigmaVar  <- c(Total = 0.388, `Unbound non-J101` = 0.404, `Unbound J101` = 0.327)
conv <- dplyr::bind_rows(
  tibble::tibble(
    Row = names(omegaVar), Block = "OMEGA", Variance = omegaVar,
    `Printed CV%` = c(53.2, 333, 24.1, 205, 24.1, 79.0),
    `sqrt(exp(v)-1)` = round(100 * sqrt(exp(omegaVar) - 1), 1),
    `sqrt(v)` = round(100 * sqrt(omegaVar), 1)
  ),
  tibble::tibble(
    Row = names(sigmaVar), Block = "SIGMA", Variance = sigmaVar,
    `Printed CV%` = c(62.3, 63.6, 57.2),
    `sqrt(exp(v)-1)` = round(100 * sqrt(exp(sigmaVar) - 1), 1),
    `sqrt(v)` = round(100 * sqrt(sigmaVar), 1)
  )
)
knitr::kable(conv, caption = "The printed CV% column identifies which scale each block uses.")

# OMEGA: sqrt(exp(v)-1) reproduces the printed CV% to the printed precision.
stopifnot(max(abs(100 * sqrt(exp(omegaVar) - 1) -
                  c(53.2, 333, 24.1, 205, 24.1, 79.0)) /
              c(53.2, 333, 24.1, 205, 24.1, 79.0)) < 0.005)
# SIGMA: sqrt(v) reproduces the printed CV%.
stopifnot(max(abs(100 * sqrt(sigmaVar) - c(62.3, 63.6, 57.2))) < 0.15)
# And the printed correlations follow from the printed covariances.
stopifnot(
  abs(0.613 / sqrt(0.249 * 2.49) - 0.778) < 0.001,
  abs(0.0629 / sqrt(0.0563 * 1.65) - 0.207) < 0.001
)
```

# Structural checks on typical values

`rxode2::zeroRe()` removes the random effects so that the following identities
can be asserted tightly; the stochastic cohort comes later.

```{r typical-model}
modT <- rxode2::zeroRe(mod)

# Observation rows must nominate one of the two endpoints; with exactly two
# outputs the idiom is dvid (1 = Cc total, 2 = Cu unbound).
buildEvents <- function(dose, nDose, grid, covs) {
  dosing <- data.frame(
    id = 1L, time = seq(0, by = 24, length.out = nDose), amt = dose,
    evid = 1L, rate = -2, cmt = "depot", dvid = NA_integer_
  )
  obs <- do.call(rbind, lapply(1:2, function(d) data.frame(
    id = 1L, time = grid, amt = NA_real_, evid = 0L, rate = NA_real_,
    cmt = NA_character_, dvid = as.integer(d)
  )))
  ev <- dplyr::arrange(rbind(dosing, obs), time, dvid)
  cbind(ev, covs[rep(1, nrow(ev)), , drop = FALSE])
}

DOSE  <- 4e5           # nmol once daily
TAU   <- 24
grid  <- seq(24 * 29, 24 * 30, by = 0.05)
ssRef <- rxode2::rxSolve(modT, buildEvents(DOSE, 30, grid, refCov),
                         returnType = "data.frame") |>
  dplyr::filter(time >= 24 * 29) |>
  dplyr::distinct(time, .keep_all = TRUE) |>
  dplyr::arrange(time)

trapz <- function(x, y) sum(diff(x) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)
aucU  <- trapz(ssRef$time, ssRef$Cu)
aucT  <- trapz(ssRef$time, ssRef$Cc)
```

## Mass balance: unbound exposure is exactly dose divided by clearance

At steady state the amount eliminated per interval must equal the amount
absorbed. Because elimination is `CL * Cu` with a constant `CL`, this pins
`AUCu,ss` to `dose / CL` exactly - a check that would fail immediately if the
binding algebra, the bioavailability, or the compartment bookkeeping were
wrong.

```{r mass-balance}
CL <- 520
c(`AUCu,ss simulated` = aucU, `dose / CL` = DOSE / CL,
  ratio = aucU / (DOSE / CL))
stopifnot(abs(aucU / (DOSE / CL) - 1) < 1e-3)
```

## Free fraction at low concentration

At `Cu << kd` the binding equation collapses to `Cu = Ctot / (1 + bmax/kd)`,
so the reference individual's low-concentration unbound fraction is fixed by
Table 2 alone at `1 / (1 + 8280/221)`.

```{r free-fraction}
# Probe the limit with a dose 10000-fold below the reference dose, so that
# Cu << kd and the quadratic collapses to its linear limit.
probe <- rxode2::rxSolve(modT, buildEvents(DOSE / 1e4, 30, grid, refCov),
                         returnType = "data.frame") |>
  dplyr::filter(time >= 24 * 29) |>
  dplyr::distinct(time, .keep_all = TRUE)
probeTrough <- probe[which.min(probe$Cc), ]

trough <- ssRef[which.min(ssRef$Cc), ]
c(`fu in the low-concentration limit, %` = 100 * probeTrough$Cu / probeTrough$Cc,
  `1/(1 + 8280/221), %`                  = 100 / (1 + 8280 / 221),
  `fu at the reference-dose trough, %`   = 100 * trough$Cu / trough$Cc,
  `fu at the reference-dose peak, %`     = 100 * ssRef$Cu[which.max(ssRef$Cc)] /
                                                 ssRef$Cc[which.max(ssRef$Cc)])
stopifnot(
  abs(100 * probeTrough$Cu / probeTrough$Cc - 100 / (1 + 8280 / 221)) < 0.05
)
```

The unbound fraction is *not* constant across the dosing interval at the
reference dose: it rises from trough to peak because the AAG binding sites
partially saturate as unbound concentrations approach `kd = 221 nmol/L`. That
concentration dependence is the whole reason the paper carries a saturable
binding sub-model rather than a fixed `fu`.

```{r free-fraction-monotone}
stopifnot(
  # fu increases with total concentration, everywhere in the interval.
  stats::cor(ssRef$Cc, ssRef$Cu / ssRef$Cc, method = "spearman") > 0.99,
  # and always sits at or above the low-concentration limit.
  min(ssRef$Cu / ssRef$Cc) >= 1 / (1 + 8280 / 221) - 1e-6
)
```

## Half-life and time to steady state

Three-compartment disposition gives this model two very different time
constants, and it is worth separating them because only one of them controls
the once-daily regimen the paper analyses.

The **effective** (mean-residence) time constant is set by the unbound-basis
steady-state volume, `V1 * (1 + bmax/kd) + V2 + V3`, over `CL`. The
**terminal** slope is set by the slowest disposition eigenvalue and is much
longer, but it carries only a small share of the area.

```{r half-life}
washGrid <- seq(0, 2000, by = 2)
wash <- rxode2::rxSolve(modT, buildEvents(DOSE, 1, washGrid, refCov),
                        returnType = "data.frame") |>
  dplyr::distinct(time, .keep_all = TRUE) |>
  dplyr::filter(Cu > 0)
tailFit <- dplyr::filter(wash, time >= 400)
tHalf <- log(2) / -stats::coef(stats::lm(log(Cu) ~ time, data = tailFit))[["time"]]
vssU  <- 42.5 * (1 + 8280 / 221) + 3670 + 2950

c(`terminal half-life, h`                 = tHalf,
  `unbound-basis Vss, L`                  = vssU,
  `effective half-life ln2 * Vss / CL, h` = log(2) * vssU / CL)
stopifnot(tHalf > 40, tHalf < 110)
```

What matters clinically is that once-daily dosing reaches steady state well
inside a treatment cycle, which is the premise of every steady-state exposure
metric the paper computes.

```{r accumulation}
troughTimes <- seq(24, 24 * 30, by = 24) - 0.01
accum <- rxode2::rxSolve(modT, buildEvents(DOSE, 30, troughTimes, refCov),
                         returnType = "data.frame") |>
  dplyr::distinct(time, .keep_all = TRUE) |>
  dplyr::arrange(time)
ssTrough <- utils::tail(accum$Cu, 1)
approach <- c(`day 3` = accum$Cu[3], `day 7` = accum$Cu[7],
              `day 14` = accum$Cu[14], `day 21` = accum$Cu[21]) / ssTrough
round(approach, 3)
stopifnot(approach[["day 7"]] > 0.90, approach[["day 14"]] > 0.97)
```

This is also the discriminator between the two readings of Figure 1. Had the
peripheral compartments equilibrated with the *total* rather than the unbound
concentration, the clearance seen by the total-drug mass balance would have
been reduced by the free fraction of about 2.6%, inflating the accumulation
time constant roughly 38-fold - so steady state would take months rather than
about a week, and the paper's steady-state exposure analysis under once-daily
dosing would be incoherent. The unbound-driven reading is the one adopted here.

# Replicating the paper's central finding: AAG moves total but not unbound exposure

Inoue 2025 concludes that "binding of valemetostat to AAG in the plasma
influenced total valemetostat exposure, but had little effect on unbound
valemetostat exposure" (Results), and that unbound exposure is the clinically
relevant metric. In the model this is not an empirical near-miss - it is an
algebraic identity created by giving CL and F1 the *same* AAG exponent.

```{r aag-sweep}
aagLevels <- c(50, 75, 100, 150, 200, 250)
aagSweep <- do.call(rbind, lapply(aagLevels, function(a) {
  cv <- refCov; cv$AAG <- a
  s <- rxode2::rxSolve(modT, buildEvents(DOSE, 30, grid, cv),
                       returnType = "data.frame") |>
    dplyr::filter(time >= 24 * 29) |>
    dplyr::distinct(time, .keep_all = TRUE) |>
    dplyr::arrange(time)
  data.frame(AAG = a, AUCu = trapz(s$time, s$Cu), AUCt = trapz(s$time, s$Cc))
}))
aagSweep <- aagSweep |>
  dplyr::mutate(
    `Unbound AUCss ratio` = AUCu / AUCu[AAG == 100],
    `Total AUCss ratio`   = AUCt / AUCt[AAG == 100]
  )
knitr::kable(
  aagSweep |> dplyr::select(AAG, `Unbound AUCss ratio`, `Total AUCss ratio`),
  digits = 4,
  caption = "AAG (mg/dL) versus steady-state exposure, relative to the reference AAG of 100 mg/dL."
)

# Unbound exposure is INVARIANT to AAG (exact, by construction).
stopifnot(max(abs(aagSweep$`Unbound AUCss ratio` - 1)) < 1e-3)
# Total exposure rises monotonically and substantially with AAG.
stopifnot(
  all(diff(aagSweep$`Total AUCss ratio`) > 0),
  aagSweep$`Total AUCss ratio`[aagSweep$AAG == 250] > 1.5
)
```

```{r aag-plot, fig.cap = "Replicates the mechanism behind Inoue 2025 Figure 2: AAG shifts total valemetostat exposure while leaving unbound exposure untouched."}
aagSweep |>
  tidyr::pivot_longer(c(`Unbound AUCss ratio`, `Total AUCss ratio`),
                      names_to = "Analyte", values_to = "Ratio") |>
  ggplot2::ggplot(ggplot2::aes(AAG, Ratio, colour = Analyte)) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::geom_point(size = 2) +
  ggplot2::geom_hline(yintercept = 1, linetype = "dashed") +
  ggplot2::labs(x = "Baseline AAG (mg/dL)",
                y = "Steady-state AUC relative to AAG = 100 mg/dL",
                colour = NULL) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "top")
```

# Covariate effects on unbound exposure reproduce Table 2

Each categorical covariate multiplies CL by its Table 2 ratio, so unbound
exposure must move by the *reciprocal* of that ratio. Simulating one covariate
at a time from the reference individual recovers exactly that.

```{r covariate-forest}
catCovs <- tibble::tribble(
  ~Covariate,                      ~Column,             ~`Table 2 CL ratio`,
  "ATLL vs PTCL",                  "TUMTP_ATLL",        0.828,
  "Other NHL vs PTCL",             "DIS_BCELLNHL",      0.847,
  "Non-patient vs PTCL",           "DIS_HEALTHY",       0.942,
  "Female vs male",                "SEXF",              1.06,
  "Asian non-Japanese vs Japanese","RACE_ASIAN_OTH",    1.25,
  "White vs Asian Japanese",       "RACE_WHITE",        0.967,
  "Other race vs Asian Japanese",  "RACE_OTHER",        0.816,
  "Mild hepatic impairment",       "HEPIMP_MILD",       0.892,
  "Moderate hepatic impairment",   "HEPIMP_MOD",        0.752
)
covResult <- do.call(rbind, lapply(seq_len(nrow(catCovs)), function(i) {
  cv <- refCov; cv[[catCovs$Column[i]]] <- 1
  s <- rxode2::rxSolve(modT, buildEvents(DOSE, 30, grid, cv),
                       returnType = "data.frame") |>
    dplyr::filter(time >= 24 * 29) |>
    dplyr::distinct(time, .keep_all = TRUE) |>
    dplyr::arrange(time)
  data.frame(Covariate = catCovs$Covariate[i],
             `Simulated unbound AUCss ratio` = trapz(s$time, s$Cu) / aucU,
             check.names = FALSE)
}))
covResult <- covResult |>
  dplyr::mutate(`Expected (1 / Table 2 ratio)` = 1 / catCovs$`Table 2 CL ratio`,
                `Relative error` = abs(`Simulated unbound AUCss ratio` /
                                         `Expected (1 / Table 2 ratio)` - 1))
knitr::kable(covResult, digits = 4,
             caption = "Categorical covariate effects on unbound steady-state exposure.")
stopifnot(max(covResult$`Relative error`) < 1e-3)
```

The paper independently quantifies the moderate-hepatic-impairment effect from
its post hoc simulations as a **29% (95% CI 1%-66%)** increase in unbound
AUCss. The structural model gives `1 / 0.752 = 1.330`, i.e. +33%. These are not
the same estimator - the paper's number is a cohort-average over the real
covariate distribution of the 12 moderately-impaired participants, whereas the
model ratio holds one covariate at a time from the reference individual - so
this is recorded as a consistency observation, not asserted.

```{r hepatic-note}
c(`structural ratio` = 1 / 0.752,
  `paper post hoc, unbound AUCss` = 1.29)
```

# Saturable binding makes total exposure less than dose-proportional

Unbound clearance is linear, so unbound exposure is exactly dose-proportional.
Total exposure is not: as unbound concentrations climb toward `kd`, the AAG
binding sites saturate and a smaller fraction of the drug is carried bound.

```{r dose-proportionality}
doseLadder <- DOSE * c(0.25, 0.5, 1, 2, 4)
dp <- do.call(rbind, lapply(doseLadder, function(d) {
  s <- rxode2::rxSolve(modT, buildEvents(d, 30, grid, refCov),
                       returnType = "data.frame") |>
    dplyr::filter(time >= 24 * 29) |>
    dplyr::distinct(time, .keep_all = TRUE) |>
    dplyr::arrange(time)
  data.frame(`Dose (nmol)` = d, AUCu = trapz(s$time, s$Cu),
             AUCt = trapz(s$time, s$Cc), check.names = FALSE)
}))
dp <- dp |>
  dplyr::mutate(`Dose ratio` = `Dose (nmol)` / DOSE,
                `Unbound AUC ratio` = AUCu / AUCu[`Dose (nmol)` == DOSE],
                `Total AUC ratio`   = AUCt / AUCt[`Dose (nmol)` == DOSE])
knitr::kable(dp |> dplyr::select(`Dose (nmol)`, `Dose ratio`,
                                 `Unbound AUC ratio`, `Total AUC ratio`),
             digits = 4, caption = "Dose proportionality of the two analytes.")

# Unbound exposure is exactly dose-proportional (linear unbound clearance).
stopifnot(max(abs(dp$`Unbound AUC ratio` / dp$`Dose ratio` - 1)) < 1e-3)
# Total exposure is strictly LESS than dose-proportional above the reference
# dose, and strictly MORE below it: the signature of saturable binding.
stopifnot(
  all(dp$`Total AUC ratio`[dp$`Dose ratio` > 1] < dp$`Dose ratio`[dp$`Dose ratio` > 1]),
  all(dp$`Total AUC ratio`[dp$`Dose ratio` < 1] > dp$`Dose ratio`[dp$`Dose ratio` < 1])
)
```

# Virtual cohort and non-compartmental analysis

A 150-participant cohort is drawn to match the PPK analysis set demographics of
Table 1, and NCA is run separately on each analyte with `PKNCA`.

```{r cohort}
set.seed(20251210)
rxode2::rxSetSeed(20251210)
nSub <- 150L

cohort <- data.frame(
  id  = seq_len(nSub),
  WT  = pmax(35, stats::rnorm(nSub, 69.3, 15.9)),
  AGE = pmin(90, pmax(20, stats::rnorm(nSub, 59.8, 17.2))),
  CRCL = pmax(20, stats::rnorm(nSub, 89.2, 35.5)),
  AAG = pmax(20, stats::rnorm(nSub, 113, 62.3)),
  SEXF = stats::rbinom(nSub, 1, 0.357),
  STUDY_J101 = 0, CONMED_PGP_INH = 0, CONMED_CYP3A4_INH = 0
)
# Disease type: PTCL reference (55.6%), ATLL 17.8%, other NHL 5.6%, non-patient 21.1%.
dz <- sample(c("PTCL", "ATLL", "otherNHL", "nonpatient"), nSub, replace = TRUE,
             prob = c(0.556, 0.178, 0.056, 0.211))
cohort$TUMTP_ATLL   <- as.integer(dz == "ATLL")
cohort$DIS_BCELLNHL <- as.integer(dz == "otherNHL")
cohort$DIS_HEALTHY  <- as.integer(dz == "nonpatient")
# Race/country: Asian Japanese reference (41.8%).
rc <- sample(c("AsianJP", "AsianNonJP", "White", "Other"), nSub, replace = TRUE,
             prob = c(0.418, 0.061, 0.392, 0.129))
cohort$RACE_ASIAN_OTH <- as.integer(rc == "AsianNonJP")
cohort$RACE_WHITE     <- as.integer(rc == "White")
cohort$RACE_OTHER     <- as.integer(rc == "Other")
# Hepatic function: normal reference (74.9%).
hf <- sample(c("normal", "mild", "moderate"), nSub, replace = TRUE,
             prob = c(0.749, 0.211, 0.035))
cohort$HEPIMP_MILD <- as.integer(hf == "mild")
cohort$HEPIMP_MOD  <- as.integer(hf == "moderate")

summary(cohort[, c("WT", "AGE", "CRCL", "AAG")])
```

```{r cohort-sim}
ncaGrid <- c(0, 0.25, 0.5, 1, 1.5, 2, 3, 4, 6, 8, 10, 12, 16, 20, 24)
dosing <- do.call(rbind, lapply(cohort$id, function(i) data.frame(
  id = i, time = seq(0, by = 24, length.out = 30), amt = DOSE,
  evid = 1L, rate = -2, cmt = "depot", dvid = NA_integer_
)))
obsRows <- do.call(rbind, lapply(cohort$id, function(i)
  do.call(rbind, lapply(1:2, function(d) data.frame(
    id = i, time = 24 * 29 + ncaGrid, amt = NA_real_, evid = 0L,
    rate = NA_real_, cmt = NA_character_, dvid = as.integer(d)
  ))))
)
evAll <- dplyr::arrange(rbind(dosing, obsRows), id, time, dvid)
evAll <- dplyr::left_join(evAll, cohort, by = "id")

simCohort <- rxode2::rxSolve(mod, evAll, returnType = "data.frame")
simSS <- simCohort |>
  dplyr::filter(time >= 24 * 29) |>
  dplyr::distinct(id, time, .keep_all = TRUE) |>
  dplyr::mutate(time = time - 24 * 29) |>
  dplyr::arrange(id, time)
nrow(simSS)
```

```{r pknca}
concU <- simSS |> dplyr::filter(!is.na(Cu)) |>
  dplyr::mutate(analyte = "Unbound", conc = Cu)
concT <- simSS |> dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(analyte = "Total", conc = Cc)

doseFrame <- cohort |> dplyr::transmute(id, time = 0, dose = DOSE)

runNca <- function(cd) {
  oConc <- PKNCA::PKNCAconc(as.data.frame(cd), conc ~ time | id)
  oDose <- PKNCA::PKNCAdose(as.data.frame(doseFrame), dose ~ time | id)
  intervals <- data.frame(start = 0, end = 24,
                          cmax = TRUE, tmax = TRUE, auclast = TRUE, cav = TRUE)
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(oConc, oDose, intervals = intervals))
  as.data.frame(res)
}
ncaU <- runNca(concU)
ncaT <- runNca(concT)

ncaSummary <- dplyr::bind_rows(
  dplyr::mutate(ncaU, Analyte = "Unbound"),
  dplyr::mutate(ncaT, Analyte = "Total")
) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "cav")) |>
  dplyr::group_by(Analyte, PPTESTCD) |>
  dplyr::summarise(Median = stats::median(PPORRES),
                   `5th pct` = stats::quantile(PPORRES, 0.05),
                   `95th pct` = stats::quantile(PPORRES, 0.95),
                   .groups = "drop") |>
  dplyr::rename(`NCA parameter` = PPTESTCD)
knitr::kable(ncaSummary, digits = 4,
             caption = "Steady-state NCA over the 24 h dosing interval, by analyte (nmol/L and nmol*h/L; tmax in h).")
```

## NCA cross-check against the structural identity

The cohort-median `cav` for the unbound analyte must sit close to
`dose / (CL * 24)` scaled by the cohort's clearance distribution. Because
clearance carries a 53% CV and weight enters allometrically, the *median* is
the right statistic to compare - the cohort extremes are not reproducible
across rxode2 builds and are deliberately not asserted.

```{r nca-check}
cavU <- ncaU |> dplyr::filter(PPTESTCD == "cav") |> dplyr::pull(PPORRES)
refCav <- DOSE / (CL * 24)
c(`median unbound cav` = stats::median(cavU),
  `reference-individual dose/(CL*24)` = refCav,
  ratio = stats::median(cavU) / refCav)

stopifnot(
  # Centre: a mis-transcribed clearance, dose or exponent moves the whole
  # distribution by tens of percent and breaks this immediately.
  abs(stats::median(cavU) / refCav - 1) < 0.30,
  # Envelope: robust to which subjects land in the tails.
  stats::quantile(cavU, 0.9) / refCav < 4,
  !anyNA(cavU)
)
```

```{r profile-plot, fig.cap = "Simulated steady-state profiles over one 24 h dosing interval (150 participants). Total valemetostat sits far above unbound; the ratio is concentration-dependent because the AAG binding sites partially saturate over the interval."}
simSS |>
  dplyr::select(id, time, Total = Cc, Unbound = Cu) |>
  tidyr::pivot_longer(c(Total, Unbound), names_to = "Analyte", values_to = "conc") |>
  dplyr::group_by(Analyte, time) |>
  dplyr::summarise(median = stats::median(conc),
                   lo = stats::quantile(conc, 0.05),
                   hi = stats::quantile(conc, 0.95), .groups = "drop") |>
  ggplot2::ggplot(ggplot2::aes(time, median, colour = Analyte, fill = Analyte)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi), alpha = 0.2, colour = NA) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::scale_y_log10() +
  ggplot2::labs(x = "Time after dose (h)", y = "Concentration (nmol/L, log scale)",
                colour = NULL, fill = NULL) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "top")
```

# The J101 unbound assay adjustment

Unbound observations from the DS3201-A-J101 study are rescaled by the estimated
factor 0.638 and carry their own residual error. Total observations are
unaffected.

```{r assay}
refJ101 <- refCov; refJ101$STUDY_J101 <- 1
sJ <- rxode2::rxSolve(modT, buildEvents(DOSE, 30, grid, refJ101),
                      returnType = "data.frame") |>
  dplyr::filter(time >= 24 * 29) |>
  dplyr::distinct(time, .keep_all = TRUE) |>
  dplyr::arrange(time)
c(`unbound AUC ratio, J101 vs other studies` = trapz(sJ$time, sJ$Cu) / aucU,
  `Table 2 ASSAY factor` = 0.638,
  `total AUC ratio (must be 1)` = trapz(sJ$time, sJ$Cc) / aucT)
stopifnot(
  abs(trapz(sJ$time, sJ$Cu) / aucU - 0.638) < 1e-3,
  abs(trapz(sJ$time, sJ$Cc) / aucT - 1)     < 1e-6
)
```

# Comedication effects on bioavailability

The paper groups the comedication categories as none / P-gp inhibitor / CYP3A
inhibitor with or without a P-gp inhibitor. The two arms are mutually
exclusive, so a participant on both belongs to the CYP3A arm - the model file
enforces that precedence explicitly rather than multiplying both factors.

```{r comed}
comedCases <- list(
  `P-gp inhibitor only`        = c(CONMED_PGP_INH = 1, CONMED_CYP3A4_INH = 0),
  `CYP3A inhibitor only`       = c(CONMED_PGP_INH = 0, CONMED_CYP3A4_INH = 1),
  `Both (CYP3A arm applies)`   = c(CONMED_PGP_INH = 1, CONMED_CYP3A4_INH = 1)
)
comedRes <- do.call(rbind, lapply(names(comedCases), function(nm) {
  cv <- refCov
  cv$CONMED_PGP_INH    <- comedCases[[nm]][["CONMED_PGP_INH"]]
  cv$CONMED_CYP3A4_INH <- comedCases[[nm]][["CONMED_CYP3A4_INH"]]
  s <- rxode2::rxSolve(modT, buildEvents(DOSE, 30, grid, cv),
                       returnType = "data.frame") |>
    dplyr::filter(time >= 24 * 29) |>
    dplyr::distinct(time, .keep_all = TRUE) |>
    dplyr::arrange(time)
  data.frame(Scenario = nm,
             `Unbound AUCss ratio` = trapz(s$time, s$Cu) / aucU,
             check.names = FALSE)
}))
comedRes$`Table 2 F1 ratio` <- c(1.29, 1.23, 1.23)
knitr::kable(comedRes, digits = 4,
             caption = "Comedication effects on unbound steady-state exposure.")
stopifnot(max(abs(comedRes$`Unbound AUCss ratio` /
                    comedRes$`Table 2 F1 ratio` - 1)) < 1e-3)
```

# Assumptions and deviations

* **Molar units.** Table 2 prints `KD` and `RMAX` in nmol/L and the paper gives
  no molar mass for valemetostat, so the model is expressed on a molar scale
  (`dosing = "nmol"`, `concentration = "nmol/L"`). A user who wants to dose in
  milligrams must supply the molar mass themselves. No molar mass was imported
  from outside the source. The doses used above are nominal nmol values chosen
  to exercise the model, not conversions of the 200 mg clinical dose. Note that
  the *unbound* predictions are unaffected by this choice: the identity
  `Cu,ss,avg = dose rate / CL` holds in any consistent unit, and applied in
  milligrams it gives 16.0 ng/mL for 200 mg once daily, bracketing the paper's
  own reference unbound Cavg values of 13.9 and 18.1 ng/mL.
* **Distribution is driven by unbound concentration.** Figure 1 draws `CL`
  leaving the "Unbound" species and the peripheral compartments attached to the
  central box; the paper does not write the differential equations out. Driving
  `Q2` and `Q3` from the unbound concentration is the reading adopted here. It
  is not a free choice: the alternative (distribution driven by total
  concentration) would shrink the clearance seen by the total-drug mass balance
  by the free fraction of about 2.6%, inflating the time to steady state some
  38-fold - months rather than about a week - which contradicts the paper's own
  steady-state exposure analysis under once-daily dosing. The accumulation
  check above is the discriminator.
* **Continuous covariates enter as power terms; categorical covariates as
  multiplicative ratios.** Table 2 does not state the functional form. The
  split is inferred from the estimates themselves: the continuous coefficients
  are centred on zero (age -0.205, CrCl -0.0107) and sit in the same block as
  the two body-weight allometric exponents, which are unambiguously exponents;
  the categorical coefficients are centred on one (ATLL 0.828, female 1.06).
* **The AAG effect on CL and F1 is one parameter, not two.** Table 2 prints
  `CL/F ~ AAG` and `F1 ~ AAG` as separate rows with the identical estimate
  (0.336) and the identical RSE (18.6%), and the Results text says AAG was
  included "by estimating a common effect on both CL/F and F1". It is encoded
  as a single `e_aag_cl` parameter used in both places. The `CL/F ~ AAG` row is
  additionally printed with an asymmetric 95% CI (`-0.165-0.567`) while the
  `F1 ~ AAG` row prints `0.165-0.567`; the two rows describe one parameter, and
  the point estimate used here is unaffected either way.
* **"Other NHL" is mapped to `DIS_BCELLNHL`.** The paper labels this population
  stratum "other NHL" without specifying histology. The only contributing study
  is DS3201-A-J101, which the Methods describe as enrolling "R/R non-Hodgkin
  lymphoma (NHL), including B-cell lymphomas, ATLL, and PTCL", so the non-ATLL
  non-PTCL remainder is the B-cell lymphoma group. A future paper whose
  residual NHL stratum is explicitly non-B-cell should not reuse this mapping.
* **Bootstrap columns not used.** Table 2's `Median` and `95% CI` columns are
  non-parametric bootstrap summaries (n = 462) and differ from the point
  estimates in places (for example `V1/F` 42.5 versus a bootstrap median of
  31.3). The `Estimate` column is used throughout, per the standing convention
  of taking the reported final estimates.
* **Residual-error parameter names.** The model carries three residual-error
  parameters because the paper estimates three SIGMAs: `propSd` (total),
  `propSd_Cu` (unbound outside J101) and `propSd_Cu_j101` (unbound within
  J101). The third is a study-conditional variant of the second rather than a
  fourth endpoint, and is selected inside `model()` by the `STUDY_J101`
  indicator.
* **No published NCA table to compare against.** Inoue 2025 reports no
  non-compartmental summary of observed concentrations, so the NCA section
  above validates internal consistency (unbound `cav` against
  `dose / (CL * 24)`) rather than reproducing a published table. The paper's
  own visual predictive checks (Figure S2) and goodness-of-fit plots (Figures
  S3-S4) are graphical and are not digitised here.
* **Companion exposure-response models.** The paper additionally reports one
  exposure-efficacy and six exposure-safety logistic-regression models
  (Tables S4-S10) that consume the unbound average concentration this PK model
  predicts. They are packaged separately as the `Inoue_2025_valemetostat_*`
  family and documented in this same vignette when available.

# Session information

```{r session}
sessionInfo()
```
