---
title: "Piperacillin + tazobactam in late elderly patients with pneumonia (Ishihara 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Piperacillin + tazobactam in late elderly patients with pneumonia (Ishihara 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Ishihara 2020 fitted piperacillin and tazobactam **separately** to 100 plasma
concentrations each from the same 18 patients (sections 2.2.1 and 2.2.2),
reporting the final estimates in two tables (Table 2 for piperacillin, Table 3
for tazobactam). The two fits are packaged as two independent model files and
described together here.

```{r models}
mod_pip <- readModelDb("Ishihara_2020_piperacillin")
mod_taz <- readModelDb("Ishihara_2020_tazobactam")
ui_pip <- rxode2::rxode(mod_pip)
ui_taz <- rxode2::rxode(mod_taz)
```

* Citation: `r ui_pip$reference`
* Piperacillin model: `r ui_pip$description`
* Tazobactam model: `r ui_taz$description`
* Article: <https://doi.org/10.3390/antibiotics9030113> (open access)

# Population

Eighteen Japanese inpatients over 75 years old ("late elderly") with pneumonia
were enrolled at the Department of Clinical Oncology and Respiratory Medicine,
Shimane University Hospital (section 4.1). Patients with serious heart, liver
or renal failure, strongly suspected atypical pneumonia, or beta-lactam allergy
were excluded.

Table 1: 14 men and 4 women; age 86.5 +/- 6.0 years (range 75-101); height
154.1 +/- 7.8 cm; body weight 45.5 +/- 10.0 kg (range 32.0-68.7); body mass
index 19.1 +/- 3.5; serum creatinine 0.91 +/- 0.31 mg/dL (range 0.60-1.55);
Cockcroft-Gault creatinine clearance 38.0 +/- 11.1 mL/min (range 21.5-59.1;
median 37.4, Table 2 footnote); serum albumin 2.9 +/- 0.6 g/dL.

Piperacillin/tazobactam 4/0.5 g was infused intravenously over 1 h three times
daily, reduced to 2/0.25 g in patients with an eGFR below 50 mL/min
(section 4.2). Blood was drawn at 0, 1, 1.5, 2, 3 and 5 h after the start of an
infusion and both analytes were assayed by HPLC-UV (quantification range
0.5-1000 ug/mL).

The same information is available programmatically as
`readModelDb("Ishihara_2020_piperacillin")()$population`.

# Source trace

Every `ini()` entry in `inst/modeldb/specificDrugs/Ishihara_2020_piperacillin.R`
and `inst/modeldb/specificDrugs/Ishihara_2020_tazobactam.R` carries an in-file
comment naming its source location. They are collected here for review.

| Equation / parameter | Piperacillin (Table 2) | Tazobactam (Table 3) | Source location |
|---|---|---|---|
| Structural model | 2-compartment, 1-h IV infusion, first-order elimination | same | Sections 2.2.1, 2.2.2; Abstract |
| `CL = theta1 + theta2 * (CLcr - 37.4)` (L/h) | `4.58 + 0.061 * (CLcr - 37.4)` | `5.00 + 0.0587 * (CLcr - 37.4)` | Tables 2 and 3 headers; footnote (37.4 = median CLcr) |
| `lcl` (theta1, L/h) | 4.58 (SE 0.289) | 5.00 (SE 0.318) | Tables 2 / 3 |
| `e_crcl_cl` (theta2, L/h per mL/min) | 0.061 (SE 0.0293) | 0.0587 (SE 0.0298) | Tables 2 / 3 |
| `lvc` (theta3, L) | 5.39 (SE 0.969) | 6.29 (SE 1.04) | Tables 2 / 3 |
| `lq` (theta4, L/h) | 20.7 (SE 6.11) | 24.0 (SE 8.44) | Tables 2 / 3 |
| `lvp` (theta5, L) | 6.96 (SE 0.314) | 7.73 (SE 0.443) | Tables 2 / 3 |
| `etalcl` (omega^2) | 0.0705 (CV 27.0%) | 0.0715 (CV 27.2%) | Tables 2 / 3 |
| `etalvc` (omega^2) | 0.389 (CV 69.0%) | 0.547 (CV 85.3%) | Tables 2 / 3 |
| `etalq` (omega^2) | 0.311 (CV 60.4%) | 0.545 (CV 85.1%) | Tables 2 / 3 |
| IIV on Vp | fixed to 0 (omitted) | fixed to 0 (omitted) | Tables 2 / 3; sections 2.2.1, 2.2.2 |
| IIV form `theta_i = theta * exp(eta_i)` | -- | -- | Section 4.3 |
| Residual `Cobs = Cpred (1 + eps_prop) + eps_add` | -- | -- | Section 4.3 |
| `propSd` = sqrt(sigma^2 prop) | sqrt(0.000927) = 0.0304 | sqrt(0.000479) = 0.0219 | Tables 2 / 3 |
| `addSd` = sqrt(sigma^2 add), mg/L | sqrt(25.1) = 5.01 | sqrt(0.394) = 0.628 | Tables 2 / 3 |
| Unbound fraction for PK/PD (not in the model) | 0.7 (30% bound) | 0.7 (30% bound) | Sections 2.3, 4.4 |

# Virtual cohort

```{r cohort}
# set.seed() seeds R's RNG for the covariate draw below. rxode2's own
# simulation RNG is seeded separately with rxSetSeed(), which fixes the eta draw
# within one rxode2 build but not across builds or thread counts -- so every
# stochastic assertion below is written on the centre / robust quantiles of the
# cohort, and the exact gates use typical values or closed forms.
set.seed(2020)
rxode2::rxSetSeed(2020)

# Truncated normal draw: Table 1 gives mean +/- SD and range for CLcr.
rtnorm <- function(n, mean, sd, lo, hi) {
  x <- rnorm(n, mean, sd)
  bad <- x < lo | x > hi
  while (any(bad)) {
    x[bad] <- rnorm(sum(bad), mean, sd)
    bad <- x < lo | x > hi
  }
  x
}

n_sub <- 150
cohort <- data.frame(
  id = seq_len(n_sub),
  # Table 1: CLcr 38.0 +/- 11.1 mL/min, range 21.5-59.1.
  CRCL = rtnorm(n_sub, 38.0, 11.1, 21.5, 59.1)
) |>
  # Section 4.2 reduces the dose when eGFR < 50 mL/min; eGFR is not reported,
  # so the Cockcroft-Gault CLcr stands in for it here.
  mutate(
    arm = ifelse(CRCL < 50, "2.25 g q8h", "4.5 g q8h"),
    dose_pip = ifelse(CRCL < 50, 2000, 4000),
    dose_taz = ifelse(CRCL < 50, 250, 500)
  )
table(cohort$arm)
```

# Simulation

```{r simulate}
# Three 1-h infusions a day (every 8 h) for 4 days. Dense grid over the first
# interval (the paper's sampling window) and the last interval (steady state).
tau <- 8
n_dose <- 12
obs_times <- sort(unique(c(
  seq(0, tau, by = 0.1),
  seq((n_dose - 1) * tau, n_dose * tau, by = 0.1)
)))

make_events <- function(cohort, dose_col) {
  dose <- cohort |>
    tidyr::crossing(dose_no = seq_len(n_dose)) |>
    mutate(
      time = (dose_no - 1) * tau, amt = .data[[dose_col]],
      rate = .data[[dose_col]], evid = 1, cmt = "central"
    ) |>
    select(id, time, amt, rate, evid, cmt, CRCL, arm)
  obs <- cohort |>
    tidyr::crossing(time = obs_times) |>
    mutate(amt = 0, rate = 0, evid = 0, cmt = "central") |>
    select(id, time, amt, rate, evid, cmt, CRCL, arm)
  bind_rows(dose, obs) |> arrange(id, time, desc(evid))
}

ev_pip <- make_events(cohort, "dose_pip")
ev_taz <- make_events(cohort, "dose_taz")

sim_pip <- rxode2::rxSolve(mod_pip, ev_pip, keep = c("CRCL", "arm"), returnType = "data.frame")
sim_taz <- rxode2::rxSolve(mod_taz, ev_taz, keep = c("CRCL", "arm"), returnType = "data.frame")
sim <- bind_rows(
  mutate(sim_pip, drug = "Piperacillin"),
  mutate(sim_taz, drug = "Tazobactam")
)
```

# Replicate published figures

```{r fig1, fig.cap = "Simulated total plasma concentrations after the first dose (median and 5th-95th percentiles), with the paper's sampling times (1, 1.5, 2, 3 and 5 h) marked. Compare with Figure 1 of Ishihara 2020 (observed concentration-time data)."}
sim |>
  filter(time <= tau) |>
  group_by(drug, arm, time) |>
  summarise(
    med = median(Cc), lo = quantile(Cc, 0.05), hi = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, med, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, colour = NA) +
  geom_line() +
  geom_vline(xintercept = c(1, 1.5, 2, 3, 5), linetype = "dotted", colour = "grey50") +
  facet_wrap(~drug, scales = "free_y") +
  labs(x = "Time after start of first infusion (h)", y = "Total plasma concentration (mg/L)",
       colour = NULL, fill = NULL) +
  theme_bw()
```

## Figure 4 -- tazobactam probability of target attainment

The tazobactam PD target is an unbound AUC over 24 h of at least 96 ug h/mL
(section 4.4), with 30% protein binding. At steady state the 24-h AUC equals the
daily dose divided by clearance, so the target reduces to
`CL <= 0.7 * D24 / 96`. With a log-normal eta on CL, the PTA has a closed form:

`PTA = Phi( (log(0.7 * D24 / 96) - log(TVCL)) / omega_CL )`

where `TVCL = 5.00 + 0.0587 * (CLcr - 37.4)` and `omega_CL^2 = 0.0715`. Volumes
and Q play no role, so this is an exact read-out of the published clearance
model. The points below were digitised by the maintainers from Figure 4.

```{r fig4}
ini_taz <- ui_taz$theta
omega_taz <- ui_taz$omega
tvcl_taz <- function(crcl) exp(ini_taz[["lcl"]]) + ini_taz[["e_crcl_cl"]] * (crcl - 37.4)
pta_taz <- function(d24, crcl) {
  pnorm((log(0.7 * d24 / 96) - log(tvcl_taz(crcl))) / sqrt(omega_taz["etalcl", "etalcl"]))
}

# Digitised from Figure 4 (TAZ dose per administration, interval).
fig4 <- tibble::tribble(
  ~regimen,                ~d24, ~`10`, ~`20`, ~`30`, ~`40`, ~`50`, ~`60`,
  "TAZ 0.5 g q6h",         2000, 100,   100,   100,   100,   100,   100,
  "TAZ 0.5 g q8h",         1500, 100,   100,   100,   100,   100,   100,
  "TAZ 0.5 g q12h",        1000, 100,   99.5,  98,    94.8,  88.5,  78.5,
  "TAZ 0.25 g q6h",        1000, 100,   99,    98,    95.5,  89.5,  79,
  "TAZ 0.25 g q8h",         750, 97.5,  91,    83,    63.5,  44.5,  27,
  "TAZ 0.25 g q12h",        500, 67,    41,    20.5,  8,     2.5,   1
) |>
  pivot_longer(`10`:`60`, names_to = "CRCL", values_to = "pta_paper") |>
  mutate(CRCL = as.numeric(CRCL), pta_model = 100 * pta_taz(d24, CRCL),
         diff = pta_model - pta_paper)

ggplot(fig4, aes(CRCL, colour = regimen)) +
  geom_line(aes(y = pta_model)) +
  geom_point(aes(y = pta_paper)) +
  labs(x = "Creatinine clearance (mL/min)", y = "PTA of fAUC0-24 >= 96 (%)",
       colour = NULL, caption = "Lines: closed form from the packaged model; points: digitised Figure 4") +
  theme_bw()

fig4 |>
  filter(d24 <= 1000) |>
  select(regimen, CRCL, pta_paper, pta_model, diff) |>
  mutate(across(c(pta_model, diff), \(x) round(x, 1))) |>
  rename(
    "Regimen" = regimen, "CLcr (mL/min)" = CRCL, "Figure 4 PTA (%)" = pta_paper,
    "Model PTA (%)" = pta_model, "Difference (pp)" = diff
  ) |>
  knitr::kable()
```

```{r fig4-gate}
# Cross-check the closed form against a full ODE solve: steady-state 24-h AUC
# for a typical subject must equal D24 / TVCL.
typ <- rxode2::zeroRe(mod_taz)
ev_typ <- rxode2::et(amt = 250, rate = 250, ii = 8, addl = 2, ss = 1, cmt = "central") |>
  rxode2::et(seq(0, 24, by = 0.01), cmt = "central") |>
  as.data.frame() |>
  mutate(CRCL = 20)
s_typ <- rxode2::rxSolve(typ, ev_typ, addDosing = FALSE, returnType = "data.frame")
auc24 <- sum(diff(s_typ$time) * (head(s_typ$Cc, -1) + tail(s_typ$Cc, -1)) / 2)
# Deterministic: same parameters on both sides; only trapezoidal error remains.
stopifnot(abs(auc24 / (750 / tvcl_taz(20)) - 1) < 0.005)

# The regimens Figure 4 plots at 100% (0.5 g q6h / q8h; section 2.3: 'achieved
# PTA >= 90% in all patients') are 98.0-100% in the closed form; the lowest is
# 0.5 g q8h at CLcr 60. Deterministic, so a fixed bound is appropriate.
stopifnot(all(fig4$pta_model[fig4$d24 >= 1500] > 97))
# The regimens with a CLcr-dependent PTA: the published curves are digitised
# and the paper's PTA came from 1000 random subjects, so the gate is on the
# centre and a robust envelope of the paper-vs-model difference.
d_mid <- fig4$diff[fig4$d24 <= 1000]
stopifnot(abs(median(d_mid)) < 5, quantile(abs(d_mid), 0.9) < 12)
```

## Table 4 and Figure 3 -- piperacillin PK/PD breakpoints

The piperacillin PD target is an unbound concentration above the MIC for at
least 50% of the dosing interval (50% fT > MIC) at steady state, with 30%
protein binding and IIV but no residual error (section 4.4). Table 4 lists the
highest MIC with PTA >= 90% for six regimens at six CLcr levels. The paper
simulated 1000 subjects per scenario; here each scenario uses 200. The model's
breakpoints are systematically higher than Table 4; the Figure 1 check and the
Assumptions and deviations section explain why the model, not Table 4, is
taken as authoritative.

```{r table4}
crcl_levels <- c(60, 50, 40, 30, 20, 10)
regimens <- tibble::tribble(
  ~regimen,      ~dose, ~ii,
  "4.5 g q6h",   4000,  6,
  "4.5 g q8h",   4000,  8,
  "4.5 g q12h",  4000, 12,
  "2.25 g q6h",  2000,  6,
  "2.25 g q8h",  2000,  8,
  "2.25 g q12h", 2000, 12
)
n_pta <- 200
scen <- tidyr::crossing(regimens, CRCL = crcl_levels) |>
  mutate(scenario = row_number())
subj <- scen |>
  tidyr::crossing(k = seq_len(n_pta)) |>
  mutate(id = row_number())

# Steady state via ss = 1, observed across one dosing interval.
ev_ss_dose <- subj |>
  mutate(time = 0, amt = dose, rate = dose, evid = 1, ss = 1, cmt = "central")
ev_ss_obs <- subj |>
  tidyr::crossing(time = seq(0, 12, by = 0.05)) |>
  filter(time <= ii) |>
  mutate(amt = 0, rate = 0, evid = 0, ss = 0, cmt = "central")
ev_ss <- bind_rows(ev_ss_dose, ev_ss_obs) |>
  select(id, time, amt, rate, ii, evid, ss, cmt, CRCL) |>
  arrange(id, time, desc(evid))
ev_ss$ii[ev_ss$evid == 0] <- 0

sim_ss <- rxode2::rxSolve(mod_pip, ev_ss, keep = "CRCL", returnType = "data.frame") |>
  left_join(select(subj, id, scenario), by = "id")

mics <- 2^(-4:7)
# Fraction of the interval with unbound concentration above each MIC; the last
# grid point (t = tau) duplicates t = 0 at steady state and is dropped.
ft <- sim_ss |>
  group_by(id, scenario) |>
  filter(time < max(time)) |>
  reframe(mic = mics, ft = vapply(mics, \(m) mean(0.7 * Cc > m), numeric(1)))
pta_pip <- ft |>
  group_by(scenario, mic) |>
  summarise(pta = mean(ft >= 0.5), .groups = "drop") |>
  left_join(scen, by = "scenario")
bp_model <- pta_pip |>
  group_by(regimen, CRCL) |>
  summarise(bp_model = max(c(0, mic[pta >= 0.9])), .groups = "drop")

table4 <- tibble::tribble(
  ~regimen,      ~`60`, ~`50`, ~`40`, ~`30`, ~`20`, ~`10`,
  "4.5 g q6h",   8,     16,    16,    32,    32,    64,
  "4.5 g q8h",   2,     4,     8,     8,     16,    32,
  "4.5 g q12h",  0.125, 0.25,  0.5,   1,     2,     4,
  "2.25 g q6h",  8,     8,     16,    16,    16,    32,
  "2.25 g q8h",  1,     2,     8,     8,     8,     16,
  "2.25 g q12h", 0.063, 0.125, 0.25,  0.5,   1,     2
) |>
  pivot_longer(-regimen, names_to = "CRCL", values_to = "bp_paper") |>
  mutate(CRCL = as.numeric(CRCL)) |>
  left_join(bp_model, by = c("regimen", "CRCL")) |>
  # The paper prints 0.0625 as 0.063.
  mutate(dilutions = round(log2(bp_model / ifelse(bp_paper == 0.063, 0.0625, bp_paper))))

table4 |>
  mutate(cell = paste0(bp_paper, " / ", bp_model)) |>
  select(regimen, CRCL, cell) |>
  pivot_wider(names_from = CRCL, values_from = cell) |>
  rename("Regimen (paper / model, ug/mL)" = regimen) |>
  knitr::kable()
```

```{r table4-gate}
table(dilutions = table4$dilutions)
# The packaged model does NOT reproduce Table 4: its breakpoints are higher in
# every cell, by one dilution for most q6h cells and by up to five for q12h
# (see "Table 4 is not reproducible" under Assumptions and deviations: Table 4
# is also inconsistent with the paper's own observed concentrations in
# Figure 1, which the model does reproduce -- gate below). The gate pins the
# direction and size of the disagreement so that a change in the model or the
# simulation that moves it is caught. A breakpoint is a step function of a
# 200-subject PTA near the 90% line, so a single cell can flip by one dilution
# between rxode2 builds; the gate is on the share of cells and the median.
stopifnot(
  mean(table4$dilutions >= 0) >= 0.9,
  median(table4$dilutions) >= 1,
  median(table4$dilutions) <= 4
)
```

## Figure 1 -- observed concentrations

Figure 1 plots the 100 observed concentrations of each drug against time after
the start of an infusion, including pre-dose (time 0) samples taken during
repeated dosing. The maintainers read the approximate range of each sampling
time off the figure (log axes, so only to within about 10-20%). The simulated
cohort's 5th-95th percentiles at steady state should overlap these ranges, and
its median should fall inside them.

```{r fig1-gate}
fig1_ranges <- tibble::tribble(
  ~drug,          ~tad, ~obs_lo, ~obs_hi,
  "Piperacillin",    0,    5,     60,
  "Piperacillin",    1,  100,    450,
  "Piperacillin",    5,   17,    130,
  "Tazobactam",      0,    1,      8,
  "Tazobactam",      1,   10,     40,
  "Tazobactam",      5,    2,     15
)
t_last <- (n_dose - 1) * tau
fig1_sim <- sim |>
  mutate(tad = round(time - t_last, 1)) |>
  filter(tad %in% c(0, 1, 5)) |>
  group_by(drug, tad) |>
  summarise(
    sim_p05 = signif(quantile(Cc, 0.05), 3), sim_median = signif(median(Cc), 3),
    sim_p95 = signif(quantile(Cc, 0.95), 3), .groups = "drop"
  ) |>
  left_join(fig1_ranges, by = c("drug", "tad"))
fig1_sim |>
  rename(
    "Drug" = drug, "Time after start of infusion (h)" = tad,
    "Figure 1 low" = obs_lo, "Figure 1 high" = obs_hi,
    "Simulated 5th pct" = sim_p05, "Simulated median" = sim_median,
    "Simulated 95th pct" = sim_p95
  ) |>
  knitr::kable()
# Centre of the simulated cohort inside the observed range at every sampling
# time, for both drugs -- robust to which subjects land in the tails.
stopifnot(with(fig1_sim, all(sim_median > obs_lo & sim_median < obs_hi)))
```

```{r fig3, fig.cap = "Model version of Figure 3 of Ishihara 2020: PTA of 50% fT > MIC for piperacillin by CLcr, one panel per MIC. The published panels are lower; see Assumptions and deviations."}
pta_pip |>
  filter(mic %in% c(2, 4, 8, 16, 32, 64)) |>
  mutate(mic = factor(paste("MIC =", mic, "ug/mL"), levels = paste("MIC =", c(2, 4, 8, 16, 32, 64), "ug/mL"))) |>
  ggplot(aes(CRCL, 100 * pta, colour = regimen, shape = regimen)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 90, linetype = "dashed", colour = "grey50") +
  facet_wrap(~mic) +
  labs(x = "Creatinine clearance (mL/min)", y = "PTA of 50% fT > MIC (%)", colour = NULL, shape = NULL) +
  theme_bw()
```

# PKNCA validation

```{r pknca}
conc <- sim |>
  filter(!is.na(Cc)) |>
  mutate(interval = ifelse(time <= tau, "first dose", "steady state")) |>
  select(id, time, Cc, drug, arm, interval)

dose_df <- bind_rows(
  mutate(ev_pip, drug = "Piperacillin"),
  mutate(ev_taz, drug = "Tazobactam")
) |>
  filter(evid == 1, time %in% c(0, (n_dose - 1) * tau)) |>
  mutate(interval = ifelse(time == 0, "first dose", "steady state")) |>
  select(id, time, amt, drug, arm, interval)

conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ time | drug + arm + interval + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | drug + arm + interval + id)
intervals <- data.frame(
  start = c(0, (n_dose - 1) * tau),
  end = c(tau, n_dose * tau),
  interval = c("first dose", "steady state"),
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE
)
nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca$result |>
  filter(PPTESTCD %in% c("cmax", "cmin", "auclast")) |>
  group_by(drug, arm, interval, PPTESTCD) |>
  summarise(median = signif(median(PPORRES), 3), .groups = "drop") |>
  pivot_wider(names_from = PPTESTCD, values_from = median) |>
  rename(
    "Drug" = drug, "Arm" = arm, "Interval" = interval,
    "Cmax (mg/L)" = cmax, "Cmin (mg/L)" = cmin, "AUC0-8 (mg h/L)" = auclast
  ) |>
  knitr::kable()
```

## Comparison against published values

The paper reports no NCA. Its Discussion states the clearance of a patient at
the cohort mean CLcr (38.0 mL/min): 4.62 L/h for piperacillin and 5.04 L/h for
tazobactam. The steady-state AUC over one interval for a typical patient at that
CLcr must therefore equal dose / CL. The simulated side is a typical-value
(`zeroRe()`) solve at CLcr = 38.0, not a cohort median, so both sides refer to
the same reference patient.

```{r nca-compare}
typ_subj <- data.frame(id = 1:2, CRCL = 38.0, dose_pip = c(2000, 4000), dose_taz = c(250, 500),
                       arm = c("2.25 g q8h", "4.5 g q8h"))
typ_sim <- bind_rows(
  rxode2::rxSolve(rxode2::zeroRe(mod_pip), make_events(typ_subj, "dose_pip"),
                  keep = "arm", returnType = "data.frame") |> mutate(drug = "Piperacillin"),
  rxode2::rxSolve(rxode2::zeroRe(mod_taz), make_events(typ_subj, "dose_taz"),
                  keep = "arm", returnType = "data.frame") |> mutate(drug = "Tazobactam")
) |>
  filter(time >= (n_dose - 1) * tau) |>
  mutate(group = paste(drug, arm))
typ_dose <- typ_sim |>
  distinct(group, id, drug, arm) |>
  mutate(time = (n_dose - 1) * tau,
         amt = ifelse(drug == "Piperacillin", typ_subj$dose_pip[id], typ_subj$dose_taz[id]))
typ_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(typ_sim, Cc ~ time | group + id),
  PKNCA::PKNCAdose(typ_dose, amt ~ time | group + id),
  intervals = data.frame(start = (n_dose - 1) * tau, end = n_dose * tau, auclast = TRUE)
))
reference <- typ_dose |>
  mutate(auclast = amt / ifelse(drug == "Piperacillin", 4.62, 5.04)) |>
  select(group, auclast)
cmp <- nlmixr2lib::ncaComparisonTable(
  typ_nca, reference, by = "group", units = c(auclast = "mg h/L")
)
knitr::kable(cmp)
# Deterministic typical-value gate: the only differences are the rounding of
# the Discussion's 4.62 / 5.04 and trapezoidal error on a 0.1 h grid.
auc_chk <- typ_nca$result |>
  filter(PPTESTCD == "auclast") |>
  left_join(reference, by = "group")
stopifnot(all(abs(auc_chk$PPORRES / auc_chk$auclast - 1) < 0.01))
```

# Assumptions and deviations

* **Two independent models.** The paper fitted piperacillin and tazobactam
  separately, so they are packaged as two files. When simulating the
  combination, their random effects are independent (no cross-drug
  correlation was estimated).
* **Tazobactam CLcr term.** No covariate reached significance for tazobactam
  (section 2.2.2); the authors retained the linear CLcr model with the
  smallest p-value by analogy with piperacillin. It is reproduced as
  published.
* **Additive-linear clearance with a proportional eta.** Section 4.3 states
  `theta_i = theta * exp(eta_i)`; the eta is applied to the whole
  covariate-adjusted CL, following `Conil_2010_tobramycin`.
* **Variance scale.** Tables 2 and 3 print omega^2 with the matching log-normal
  CV (e.g. sqrt(exp(0.0705) - 1) = 27.0%), and sigma^2 for residual error; the
  model stores the variances for omega and the square roots for `propSd` /
  `addSd`. The combined error uses separate epsilons (variances add), matching
  `Cpred (1 + eps_prop) + eps_add`.
* **Dose-reduction rule.** The study reduced the dose when eGFR was below
  50 mL/min; eGFR was not reported, so the virtual cohort uses Cockcroft-Gault
  CLcr for that split.
* **CLcr levels outside the fitted range.** The PTA scenarios at 10, 20 and
  60 mL/min, taken from the paper, lie outside the observed CLcr range
  (21.5-59.1 mL/min); they extrapolate the linear clearance model exactly as the
  paper did.
* **Figure 4 residuals.** The model reproduces Figure 4's ordering, the
  identical curves for the equal-daily-dose regimens (0.5 g q12h and
  0.25 g q6h), and the 100% plateaus. The published curves are somewhat
  steeper in CLcr than the closed form (the paper is a few percentage points
  higher at low CLcr and lower at high CLcr). This cannot come from the
  structural model, since the closed form uses only the published CL equation
  and omega; it is most likely Monte Carlo noise in the paper's 1000-subject
  simulation plus digitisation error. Parameters were not tuned.
* **Table 4 and Figure 3 are not reproducible from Table 2.** Simulated as
  described in section 4.4 (steady state, 1-h infusion, IIV only, 30% protein
  binding, 50% fT > MIC), the packaged piperacillin model gives PK/PD
  breakpoints above Table 4 in every cell: typically one dilution for the q6h
  regimens and three to five for q12h. The maintainers checked the alternative
  reading that the paper simulated a one-compartment model with V = Vc; that
  undershoots Table 4 by two to four dilutions, so it is not the explanation
  either. Table 4 also conflicts with the paper's own data: Figure 1 shows
  pre-dose troughs of about 5-60 mg/L and 5-h concentrations of about
  17-130 mg/L, mostly after 2-g doses. Unbound piperacillin at mid-interval
  is therefore far above the 0.063-4 ug/mL q12h breakpoints of Table 4. The
  model reproduces Figure 1 (gate above), and the tazobactam PTA of Figure 4,
  which depends only on CL and omega_CL, is reproduced with a median
  difference of about -2 percentage points (at most about 9). The discrepancy is therefore attributed to the paper's
  piperacillin PTA simulation, not to the Table 2 estimates, which are
  packaged as published. Nothing was tuned. Figure 3 above is the model's
  version and differs from the published panels in the same direction.
* **Unbound fraction.** The 30% protein binding (fu = 0.7) used for the PK/PD
  analyses is applied in this vignette only; the model's `Cc` is the total
  plasma concentration that was measured and fitted.
