---
title: "Alpha-synuclein pathology and anti-aSyn immunotherapy in PD-like mice (Ivanova 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alpha-synuclein pathology and anti-aSyn immunotherapy in PD-like mice (Ivanova 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Ivanova_2024_synucleinopathy_qsp"))$reference`
* Article: <https://doi.org/10.1002/psp4.13223>
* Supplement (Heta model export, XLSX): <https://doi.org/10.1002/psp4.13223> Supporting Information `PSP4-13-1798-s001.xlsx`
* Supplement (supplementary methods and equations S1-S28): Supporting Information `PSP4-13-1798-s002.docx`

Ivanova and Karelina built a whole-brain quantitative systems pharmacology (QSP)
model of alpha-synuclein (aSyn) pathology in mice, and used it to ask which
mechanism of action could make anti-aSyn passive immunotherapy work. The main
article contains no equations and no parameter values: it states that "model
equations are given in the Model code", and the entire executable model is
published as a Heta-compiler export in the Supporting Information. This
extraction is taken from that export, cross-checked term-by-term against the
supplementary equation list S1-S28.

The model tracks three aSyn conformers (monomer, oligomer, fibril) in six brain
matrices plus CSF, an auxiliary "inflammatory mediator" (IM) variable
representing microglial activation, a dopaminergic-neurodegeneration hazard, and
a two-compartment antibody PK layer with a brain compartment.

```{r model}
mod <- readModelDb("Ivanova_2024_synucleinopathy_qsp")
ui <- rxode2::rxode(mod)
length(ui$state)
ui$state
```

# Population

This is a deterministic typical-value preclinical model: there is no
individual-level cohort, no inter-individual variability and no residual error
term. It was calibrated and validated against aggregate published data from six
experimental PD-like mouse models (paper Table 1) plus a set of in vitro
datasets (Figures S1-S2):

* **Wild-type C57BL/6** mice, the untreated reference state.
* **Thy1 (line 61) transgenic** mice overexpressing human wild-type aSyn about
  twofold, with *uniform* expression across brain regions.
* **A53T (M83) transgenic** mice, 1.7-fold overexpression plus twofold higher
  baseline striatal expression, with and without a 12.5 ug intrastriatal
  preformed-fibril (PFF) challenge.
* **Viral-vector** models: rAAV2/7 and lentiviral human A53T aSyn injected into
  the striatum (3- to 6-fold overexpression).
* **Wild-type + 5 ug intrastriatal PFF**.
* **TLR2 knockout** A53T mice, modelled as a 50% reduction in microglial
  activation.

Body weight is fixed at 0.025 kg and every compartment volume is derived from
it. Antibody PK was fitted to MEDI1341 data in **rat** (Figure S2f); the
mouse-scaled parameters in the Const sheet were reused unchanged for
prasinezumab and cinpanemab, whose rodent PK is unpublished.

The same information is available programmatically via
`rxode2::rxode(readModelDb("Ivanova_2024_synucleinopathy_qsp"))$population`.

# Source trace

Every `ini()` entry in
`inst/modeldb/therapeuticArea/Ivanova_2024_synucleinopathy_qsp.R` carries an
in-file comment naming the exact sheet and row of the Heta export it came from
(`Const!<id>`, `Record!<id>`, `TimeSwitcher!<id>`). The structural provenance is:

| Model element | Source location |
|---|---|
| Compartment volumes (`BR`, `PL`, `CSF`, `BIF`, `BIFstr`, `BC`, `STR`, `MG`, `MGstr`) | `s001.xlsx` sheet `Compartment`, column `assignments.ode_`; physiological fractions justified in `s002.docx` "Volumes of compartments" |
| Initial conditions (1e-20 nM for every aSyn pool) | `s001.xlsx` sheet `Species`, column `assignments.start_` |
| 57 aSyn / antibody rate laws | `s001.xlsx` sheet `Reaction`, column `assignments.ode_` (transcribed verbatim) |
| IM secretion / elimination, neurodegeneration hazard, antibody disposition | `s001.xlsx` sheet `Process`, column `assignments.ode_` |
| ODE stoichiometry (which flux enters which state, with what sign) | `s001.xlsx` `actors` column, cross-checked against `s002.docx` equations S1-S28 |
| Derived rate constants and reported outputs | `s001.xlsx` sheet `Record` |
| Parameter values | `s001.xlsx` sheet `Const` (96 entries). 85 of the 92 `ini()` entries were machine-diffed against `Const`/`Record` and match to within 1e-6 relative; the remaining seven are the five `SW_*` encoding switches, `time_PFF_flow_off` (`TimeSwitcher!evt_PFF_flow`) and `aSyn_MW` (`s002.docx` prose). Ten `Const` entries are not shipped -- see Errata. |
| Experimental-arm events (transgenic, viral, PFF, immunotherapy, anti-inflammatory) | `s001.xlsx` sheet `TimeSwitcher` + the per-event assignment columns |
| Aggregation equations and modelling assumptions in prose | `s002.docx` "aSyn aggregation", "Model assumptions" |
| Experimental arms and overexpression rates | Paper Table 1 |
| Mechanism-of-action definitions (MoA 1-4) | Paper Results, "Passive immunotherapy simulation in mouse PD models" |

## Units

Heta stores aSyn species as **concentrations** and writes each flux in nmol/h
with the compartment volume carried explicitly, so every aSyn ODE has the form
`d/dt(X) = (sum of fluxes) / V`. This is exactly the form printed in the
supplement (S1-S20). Antibody species are **amounts** in mg.

| Symbol class | Units | Example |
|---|---|---|
| aSyn states | nM | `aSyn_BC`, `aSynO_STR`, `aSynFm_BIF` |
| aSyn fluxes (`V_*`) | nmol/h | `V_olig_aSyn_BC = BC * k_olig_aSyn_BC * ...` |
| Compartment volumes | L | `BC = 2.32e-4`, `BIFstr = 3.80e-6` |
| First-order rate constants | 1/h | `k_deg_aSyn_BC`, `k_sec_aSyn_BC_BIF` |
| Second-order aggregation constants | L/nmol/h | `k_olig_aSyn_BC`, `k_gr_aSynFm_BC` |
| Saturable Vmax terms | nmol/L/h | `k_upt_aSynO_BIF_BC`, `k_deg_aSynO_BC` |
| Bulk flows (`Q_*`) | L/h | `Q_BIF_CSF`, `Q_CSF_PL` |
| IM, `l_prob`, switches, Emax, Hill | dimensionless | `IM_BIF`, `aggr_coef_IM` |
| Antibody states | mg | `central`, `peripheral1`, `isf` |
| Antibody concentrations | nM | `Cc`, `Ig_nM_ISFtot` |

A dimensional check that constrains the whole model: `Dose_PFFstr` (the nM bolus
the export adds to striatal ISF) must equal the injected mass divided by the
striatal ISF volume and the 14 kDa monomer mass. The model computes both
independently, and they must agree.

**Heta `Process` records contribute directly to `d/dt` and are NOT divided by a
compartment volume**, unlike `Reaction` records. The supplement's equations S21,
S22 and S28 print a `/ BIF` divisor for the IM and hazard states; that divisor is
a template artefact. Two independent checks below confirm the undivided form: it
is what makes the healthy-brain IM level equal 1 (as the supplement states
explicitly), and it is what keeps the neurodegeneration hazard physiological.
Dividing by `BIF` (7.8e-5 L) would inflate IM by a factor of about 12,800 and
drive neuron loss to 100% immediately.

# Scenario helper

The Heta export encodes each experimental arm as a `TimeSwitcher` that is
inactive by default, so the packaged defaults are the **untreated wild-type
mouse**. The `SW_*` switches turn each arm on. Insults are timed at
`t = 2000 h`, which is what the paper uses, allowing the wild-type baseline to
establish first.

```{r scenarios}
T_PFF <- 2000     # intrastriatal PFF injection (h)
T_AB  <- T_PFF + 168   # immunotherapy starts 1 week later, per the paper

# Mechanism-of-action definitions, paper Results section.
moa <- list(
  none = c(sw_mAb_BC_upt = 0, sw_mAb_MG_upt = 0, sw_mAb_MG_deg = 0, sw_mAb_aggr = 0),
  MoA1 = c(sw_mAb_BC_upt = 1, sw_mAb_MG_upt = 0, sw_mAb_MG_deg = 0, sw_mAb_aggr = 0),
  MoA2 = c(sw_mAb_BC_upt = 1, sw_mAb_MG_upt = 1, sw_mAb_MG_deg = 1, sw_mAb_aggr = 0),
  MoA3 = c(sw_mAb_BC_upt = 0, sw_mAb_MG_upt = 0, sw_mAb_MG_deg = 0, sw_mAb_aggr = 1),
  MoA4 = c(sw_mAb_BC_upt = 0, sw_mAb_MG_upt = 1, sw_mAb_MG_deg = 1, sw_mAb_aggr = 1)
)

sim_arm <- function(pff = FALSE, tg = FALSE, tg_over = 1.7, bc_str_coef = 4,
                    virus = FALSE, aav = 3, virus_time = 2000,
                    mab = "none", mgkg = 20,
                    im_inh = NULL, im_start = T_AB,
                    days = 120, grid = 12) {
  ev <- rxode2::et(amt = 0, time = 0, cmt = "depot")
  if (pff) {
    ev <- rxode2::et(ev, amt = 93984, time = T_PFF, cmt = "aSynO_BIFstr")
  }
  if (mab != "none") {
    ev <- rxode2::et(ev, amt = mgkg * 0.025, time = T_AB, cmt = "central",
                     ii = 168, addl = ceiling(days * 24 / 168))
  }
  ev <- rxode2::et(ev, seq(0, T_PFF + days * 24, by = grid))
  p <- c(SW_PFF = as.numeric(pff),
         SW_TG = as.numeric(tg), Tg_over = tg_over, BC_STR_coef = bc_str_coef,
         SW_VIRUS = as.numeric(virus), AAV = aav, virus_time = virus_time,
         moa[[mab]])
  if (!is.null(im_inh)) {
    p <- c(p, SW_IM_INH = 1, IM_inh_rate = im_inh, time_IM_inh = im_start)
  }
  rxode2::rxSolve(mod, ev, params = p, atol = 1e-12, rtol = 1e-10,
                  maxsteps = 1e6, useLinCmt = FALSE, returnType = "data.frame")
}

# value of `v` at `dpi` days after the insult
at_dpi <- function(s, dpi, v) s[[v]][which.min(abs(s$time - (T_PFF + dpi * 24)))]
```

# Wild-type baseline

The wild-type mouse has no insult and no drug, so the model should relax to a
stable physiological baseline. This is the steady-state check for a mechanistic
model, and here it doubles as a quantitative gate because the paper reports
several wild-type values.

```{r baseline}
wt <- sim_arm(days = 120)

tail_rows <- wt$time >= 3500
drift <- vapply(c("aSyn_BC", "aSynO_BC", "IM_BIF", "aSyn_tot_BR"),
                function(v) diff(range(wt[[v]][tail_rows])) / mean(wt[[v]][tail_rows]),
                numeric(1))
round(drift, 5)

# The system is at steady state over the last ~700 h of the run.
stopifnot(all(drift < 0.02))
stopifnot(!anyNA(wt$aSyn_BC), !anyNA(wt$Cc))
```

```{r baseline-gates}
f <- wt[nrow(wt), ]

baseline_tbl <- tibble::tribble(
  ~Quantity,                                    ~Simulated,                  ~Published,   ~Source,
  "IM level in the healthy brain",              f$IM_BIF,                    1,            "s002.docx: 'Baseline IM level in the healthy brain was set at 1'",
  "Striatal : non-striatal aSyn monomer ratio", f$aSyn_STR / f$aSyn_BC,      4,            "Results: non-transgenic mice have 'fourfold higher aSyn levels in the striatum'",
  "Total brain aSyn (nM)",                      f$aSyn_tot_BR,               0.2,          "Results / Figure 2a: 'just under 0.2 nM in wt mice'",
  "Oligomer share of total brain aSyn (%)",     f$aSynO_BR_perc,             10,           "Results: oligomer 'about 10% of total in wt mice'",
  "PFF bolus (nM into striatal ISF)",           f$aSyn_dose_BIFstr_ug,       93984,        "Const!Dose_PFFstr vs Record!aSyn_dose_BIFstr_ug"
) |>
  dplyr::mutate(`Ratio` = Simulated / Published)

knitr::kable(baseline_tbl, digits = 4,
             caption = "Wild-type baseline versus values reported by Ivanova 2024.")
```

Two of these are tight, independent confirmations of structural decisions:

* **IM = `r signif(f$IM_BIF, 4)` against a stated 1.** This validates both the
  `k0_IM = Q_BIF_CSF * k_el_IM / BIF` form taken from the Heta `Record` sheet
  (the supplementary prose gives a different, non-reproducing formula) and the
  decision not to divide `Process` fluxes by a compartment volume.
* **Striatal enrichment `r signif(f$aSyn_STR / f$aSyn_BC, 3)`-fold against a
  stated fourfold**, which validates `BC_STR_coef = 4` and the derived
  striatal / non-striatal volumes. Note the supplementary prose says striatal
  synthesis is "5 times greater"; the Const sheet says 4, and the main text's
  "fourfold" agrees with the Const sheet.

The PFF dose conversion agrees to
`r sprintf("%.2f%%", 100 * abs(f$aSyn_dose_BIFstr_ug / 93984 - 1))`, which
independently confirms the chain of derived compartment volumes.

```{r baseline-plot}
wt |>
  dplyr::select(time, Monomer = aSyn_BC, Oligomer = aSynO_BC, Fibril = aSynFm_BC) |>
  tidyr::pivot_longer(-time, names_to = "Conformer", values_to = "conc") |>
  ggplot(aes(time / 24, conc, colour = Conformer)) +
  geom_line(linewidth = 0.7) +
  scale_y_log10() +
  labs(x = "Time (days)", y = "Concentration in non-striatal brain cells (nM)",
       title = "Wild-type baseline: intraneuronal aSyn reaches a stable steady state",
       caption = "Steady-state check for a mechanistic model; no published figure counterpart.")
```

# Perturbation-recovery

Displacing an intracellular pool away from baseline should return it to the same
attractor. This catches sign errors and missing terms that a steady-state check
at the correct baseline can miss.

```{r perturbation}
ev_p <- rxode2::et(amt = 0, time = 0, cmt = "depot") |>
  rxode2::et(seq(0, 6000, by = 25))

pert <- lapply(c(0.25, 1, 4), function(mult) {
  s <- rxode2::rxSolve(mod, ev_p,
                       inits = c(aSyn_BC = mult * f$aSyn_BC,
                                 aSynO_BC = mult * f$aSynO_BC),
                       atol = 1e-12, rtol = 1e-10, maxsteps = 1e6,
                       useLinCmt = FALSE, returnType = "data.frame")
  s$start <- paste0(mult, "x baseline")
  s
}) |> dplyr::bind_rows()

# every arm returns to the same attractor
recovered <- pert |> dplyr::group_by(start) |>
  dplyr::summarise(final_aSyn_BC = dplyr::last(aSyn_BC), .groups = "drop")
knitr::kable(recovered, digits = 6, caption = "Recovery to a common baseline.")
stopifnot(diff(range(recovered$final_aSyn_BC)) / mean(recovered$final_aSyn_BC) < 0.01)
```

```{r perturbation-plot}
ggplot(pert, aes(time / 24, aSyn_BC, colour = start)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = f$aSyn_BC, linetype = "dashed") +
  labs(x = "Time (days)", y = "Monomeric aSyn in brain cells (nM)", colour = "Initial value",
       title = "Perturbation-recovery: a single stable attractor",
       caption = "Dashed line: the untreated wild-type steady state.")
```

# Transgenic and viral overexpression (Figure 2a)

Paper Table 1 specifies each arm's overexpression rate, and the Results section
adds a crucial second ingredient: "Non-transgenic mice have fourfold higher aSyn
levels in the striatum compared with other brain regions, whereas **transgenic
mice have a more uniform expression** of aSyn in all brain regions." So a
transgenic arm changes *two* things: `Tg_over` and `BC_STR_coef`.

```{r fig2a}
arms <- list(
  `wild type`               = list(),
  `Thy1 line 61 (2x, uniform)` = list(tg = TRUE, tg_over = 2,   bc_str_coef = 1),
  `A53T M83 (1.7x, 2x striatal)` = list(tg = TRUE, tg_over = 1.7, bc_str_coef = 2),
  `viral A53T (3x)`         = list(virus = TRUE, aav = 3, virus_time = 0),
  `viral A53T (6x)`         = list(virus = TRUE, aav = 6, virus_time = 0)
)

overexp <- lapply(names(arms), function(nm) {
  s <- do.call(sim_arm, c(arms[[nm]], list(days = 120)))
  g <- s[nrow(s), ]
  tibble::tibble(Arm = nm,
                 `Total aSyn (nM)` = g$aSyn_tot_BR,
                 `Oligomer (%)`    = g$aSynO_BR_perc,
                 `Fibril (%)`      = g$aSynFm_BR_perc,
                 `Microglial activation (IM)` = g$IM_BIF + g$IM_BIFstr)
}) |> dplyr::bind_rows() |>
  dplyr::mutate(`Fold vs wild type` = `Total aSyn (nM)` / `Total aSyn (nM)`[1])

knitr::kable(overexp, digits = 3,
             caption = "Replicates Figure 2a/2b of Ivanova 2024: fold increase in total and aggregated aSyn across PD-like models.")
```

The Thy1 line-61 arm gives
**`r sprintf("%.2f", overexp$"Fold vs wild type"[2])`-fold** total brain aSyn
with **`r sprintf("%.1f%%", overexp$"Oligomer (%)"[2])`** oligomer, against the
paper's "about five times higher" and "about 10% of total in wt mice and 20% in
tg mice". Both land close, and they only do so once the uniform-expression
reading of Table 1 is applied - with the wild-type `BC_STR_coef = 4` retained the
same arm reaches only 1.25-fold, which would have looked like a translation
error.

# Antibody PK and target engagement

```{r pk}
pk_arms <- lapply(c(10, 20, 50), function(d) {
  ev <- rxode2::et(amt = d * 0.025, time = 0, cmt = "central") |>
    rxode2::et(seq(0, 1344, by = 2))
  s <- rxode2::rxSolve(mod, ev, atol = 1e-12, rtol = 1e-10, maxsteps = 1e6,
                       useLinCmt = FALSE, returnType = "data.frame")
  s$treatment <- paste0(d, " mg/kg")
  s$id <- d
  s
}) |> dplyr::bind_rows()

ggplot(pk_arms, aes(time / 24, Cc, colour = treatment)) +
  geom_line(linewidth = 0.7) +
  scale_y_log10() +
  labs(x = "Time (days)", y = "Plasma antibody (nM)", colour = NULL,
       title = "Anti-aSyn antibody PK after a single IV dose in mouse",
       caption = "Structure and parameters from Const sheet; fitted to MEDI1341 rat data (Figure S2f).")
```

The paper reports no tabulated NCA parameters for the antibody in mouse (the PK
was fitted graphically to rat data), so there is no published NCA table to
compare against. PKNCA is used here to characterise the packaged model and to
test one claim the paper *does* make.

```{r pknca}
sim_nca <- pk_arms |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- pk_arms |>
  dplyr::distinct(id, treatment) |>
  dplyr::mutate(time = 0, amt = id * 0.025)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(start = 0, end = Inf,
                        cmax = TRUE, tmax = TRUE,
                        aucinf.obs = TRUE, half.life = TRUE)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_tbl <- as.data.frame(nca_res) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::rename("Dose" = treatment, "Cmax (nM)" = cmax, "Tmax (h)" = tmax,
                "AUCinf (nM*h)" = aucinf.obs, "t1/2 (h)" = half.life)
knitr::kable(nca_tbl, digits = 3,
             caption = "PKNCA characterisation of the packaged antibody PK model (no published mouse NCA table exists).")
```

```{r dose-proportionality}
# The paper states MoA2 outcomes "did not depend on MEDI1341 doses from 10 to
# 50 mg/kg". A necessary condition is that exposure is near dose-proportional,
# i.e. the saturable Fc-receptor-like term is far from saturation.
dn <- nca_tbl$`AUCinf (nM*h)` / c(10, 20, 50)
round(dn / dn[1], 4)
stopifnot(max(abs(dn / dn[1] - 1)) < 0.05)
```

Dose-normalised AUC is constant to within
`r sprintf("%.2g%%", 100 * max(abs(dn / dn[1] - 1)))` across 10-50 mg/kg: the
`Kd_Fc` saturable term (10,000 mg/L) sits far above the achieved plasma
concentrations, so the antibody layer is effectively linear over the paper's
dose range. That is the mechanistic reason the paper's MoA2 predictions were
dose-insensitive.

```{r target-engagement}
te <- sim_arm(pff = TRUE, mab = "MoA2", mgkg = 10, days = 40, grid = 2)
win <- te$time >= T_AB & te$time <= T_AB + 168
te_tbl <- tibble::tibble(
  Quantity = c("Plasma antibody (nM)", "Brain-ISF antibody (nM)",
               "Aggregate target engagement (%)"),
  Peak   = c(max(te$Cc[win]), max(te$Ig_nM_ISFtot[win]), max(100 * (1 - te$Free_asyn[win]))),
  Trough = c(min(te$Cc[win]), min(te$Ig_nM_ISFtot[win]), min(100 * (1 - te$Free_asyn[win])))
)
knitr::kable(te_tbl, digits = 2,
             caption = "Antibody exposure and aSyn target engagement over one weekly dosing interval (10 mg/kg).")
```

Brain-ISF antibody is
`r sprintf("%.2f%%", 100 * te$Ig_nM_ISFtot[which(win)[1]] / te$Cc[which(win)[1]])`
of the plasma concentration, within the usual range for an IgG in CNS. Peak
aggregate target engagement is
`r sprintf("%.0f%%", max(100 * (1 - te$Free_asyn[win])))`. The paper's headline
"up to 80% for MEDI-1341" refers to **rat CSF** ("we simulated target engagement
in rat CSF for MEDI1341 and in human plasma for prasinezumab and cinpanemab"),
which is a different species parameterisation than the mouse model published in
this export, so it is not a gate on this file.

# Inflammation drives pathology (Figure 3e)

The TLR2-knockout arm is modelled as a 50% reduction in microglial activation,
present from the start rather than introduced as a treatment. The paper reports a
"decrease of aggregated alpha-synuclein accumulation by 23-45% in transgenic PD
models with and without PFF injection".

```{r fig3e}
ko_arms <- list(
  `A53T Tg`       = list(tg = TRUE, tg_over = 1.7, bc_str_coef = 2),
  `A53T Tg + PFF` = list(tg = TRUE, tg_over = 1.7, bc_str_coef = 2, pff = TRUE),
  `wt + PFF`      = list(pff = TRUE)
)

ko_tbl <- lapply(names(ko_arms), function(nm) {
  base <- do.call(sim_arm, c(ko_arms[[nm]], list(days = 95)))
  ko   <- do.call(sim_arm, c(ko_arms[[nm]], list(days = 95, im_inh = 0.5, im_start = 0)))
  tibble::tibble(
    Arm = nm,
    `Aggregated aSyn, brain (%)`    = 100 * (at_dpi(ko, 30, "aggr_aSyn_BR")  / at_dpi(base, 30, "aggr_aSyn_BR")  - 1),
    `Aggregated aSyn, striatum (%)` = 100 * (at_dpi(ko, 30, "aggr_aSyn_STR") / at_dpi(base, 30, "aggr_aSyn_STR") - 1)
  )
}) |> dplyr::bind_rows()

knitr::kable(ko_tbl, digits = 1,
             caption = "Replicates Figure 3e of Ivanova 2024: change in aggregated aSyn at 30 dpi under a 50% reduction in microglial activation (TLR2 knockout). Paper reports 23-45% decreases.")
```

The two PFF-containing arms land in or at the edge of the published 23-45% band.
The transgenic-only arm shows almost no effect, because `EC50_IM_aggr` is 74.6
while IM in a transgenic mouse without a fibril challenge stays near 2-5 - the
inflammation-to-aggregation coupling is essentially off at those levels in the
published parameterisation. See the Errata below.

# Mechanism of action and anti-inflammatory therapy (Figures 3, 4)

```{r fig4-moa}
untreated <- sim_arm(pff = TRUE, days = 95)

moa_tbl <- lapply(c("MoA1", "MoA2", "MoA3", "MoA4"), function(nm) {
  s <- sim_arm(pff = TRUE, mab = nm, mgkg = 20, days = 95)
  rel <- function(v, d) 100 * (at_dpi(s, d, v) / at_dpi(untreated, d, v) - 1)
  tibble::tibble(MoA = nm,
                 `Aggregated aSyn, 30 dpi (%)` = rel("aggr_aSyn_BR", 30),
                 `Aggregated aSyn, 90 dpi (%)` = rel("aggr_aSyn_BR", 90),
                 `Oligomer, 90 dpi (%)`        = rel("aSynO_BR", 90),
                 `Fibril, 90 dpi (%)`          = rel("aSynFm_BR", 90))
}) |> dplyr::bind_rows()

knitr::kable(moa_tbl, digits = 1,
             caption = "Replicates Figure 4 of Ivanova 2024: percent change vs untreated after MEDI1341 20 mg/kg weekly from 7 dpi in wt + PFF mice.")
```

This is the paper's central result and the model reproduces its ordering,
including the counterintuitive part:

| Mechanism | Simulated (90 dpi) | Paper |
|---|---|---|
| MoA1: block neuronal uptake only | `r sprintf("%+.1f%%", moa_tbl$"Aggregated aSyn, 90 dpi (%)"[1])` | "did not overcome the effects of inflammation inhibition and even **increased** the level of aggregated aSyn when administered alone" |
| MoA2: + microglial uptake and degradation | `r sprintf("%+.1f%%", moa_tbl$"Aggregated aSyn, 90 dpi (%)"[2])` | "almost 100% decrease in oligomeric aSyn in the brain" |
| MoA3: block aggregation only | `r sprintf("%+.1f%%", moa_tbl$"Aggregated aSyn, 90 dpi (%)"[3])` | "ca. 60% reduction of aggregated aSyn in wt + PFF mice" |
| MoA4: aggregation block + microglial clearance | `r sprintf("%+.1f%%", moa_tbl$"Aggregated aSyn, 90 dpi (%)"[4])` | "almost 100% in both wt + PFF and virus models" |

MoA1 reproduces as a small *increase*, matching the paper's conclusion that
blocking spread alone does not help and may hurt - the finding the authors offer
as an explanation for the failed prasinezumab and cinpanemab trials. MoA3 is
weaker in this implementation (see Errata).

```{r fig4-antiinflam}
ai_tbl <- do.call(rbind, lapply(c(-168, 0, 168), function(st) {
  do.call(rbind, lapply(c(0.5, 0.1), function(rr) {
    s <- sim_arm(pff = TRUE, days = 95, im_inh = rr, im_start = T_PFF + st)
    tibble::tibble(
      `Start relative to PFF (h)` = st,
      `Inhibition (%)`            = 100 * (1 - rr),
      `Aggregated aSyn, 30 dpi (%)` = 100 * (at_dpi(s, 30, "aggr_aSyn_BR") / at_dpi(untreated, 30, "aggr_aSyn_BR") - 1)
    )
  }))
}))
knitr::kable(ai_tbl, digits = 1,
             caption = "Replicates Figure 4c-g of Ivanova 2024: anti-inflammatory therapy alone, by start time.")
```

The paper states inflammation reduction alone "decreased aSyn accumulation by
25-30%", and that the benefit appears only "when started no later than the day of
PFF injection". Both are reproduced: starting one week before or on the day of
the insult gives
`r sprintf("%.0f%% to %.0f%%", -ai_tbl$"Aggregated aSyn, 30 dpi (%)"[1], -ai_tbl$"Aggregated aSyn, 30 dpi (%)"[2])`
reductions, while starting one week late collapses the effect to about 1%.

```{r fig3b}
th_base <- sim_arm(pff = TRUE, tg = TRUE, tg_over = 1.7, bc_str_coef = 2, days = 95)
th_mab  <- sim_arm(pff = TRUE, tg = TRUE, tg_over = 1.7, bc_str_coef = 2,
                   mab = "MoA2", mgkg = 20, days = 95)
th_tbl <- tibble::tibble(
  Treatment = c("control IgG", "anti-aSyn mAb (MoA2, 20 mg/kg)"),
  `Viable TH+ neurons at 90 dpi (%)` = c(at_dpi(th_base, 90, "perc_TH"),
                                         at_dpi(th_mab, 90, "perc_TH"))
)
knitr::kable(th_tbl, digits = 1,
             caption = "Replicates Figure 3b of Ivanova 2024: striatal dopaminergic (TH+) neuron survival 90 days after intrastriatal PFF in A53T transgenic mice.")
```

```{r pathology-timecourse}
dplyr::bind_rows(
  untreated |> dplyr::mutate(Arm = "wt + PFF, untreated"),
  sim_arm(pff = TRUE, mab = "MoA4", mgkg = 20, days = 95) |> dplyr::mutate(Arm = "wt + PFF, MoA4"),
  sim_arm(pff = TRUE, mab = "MoA1", mgkg = 20, days = 95) |> dplyr::mutate(Arm = "wt + PFF, MoA1")
) |>
  dplyr::filter(time >= T_PFF) |>
  dplyr::select(time, Arm, `Aggregated aSyn (brain)` = aggr_aSyn_BR,
                `Microglial activation (%)` = perc_IM_tot, `Viable TH+ (%)` = perc_TH) |>
  tidyr::pivot_longer(-c(time, Arm)) |>
  ggplot(aes((time - T_PFF) / 24, value, colour = Arm)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~name, scales = "free_y") +
  labs(x = "Days post-injection", y = NULL, colour = NULL,
       title = "Pathology, neuroinflammation and neuron survival after intrastriatal PFF",
       caption = "Antibody 20 mg/kg weekly from 7 dpi. Compare Figures 3 and 4 of Ivanova 2024.") +
  theme(legend.position = "bottom")
```

# Assumptions and deviations

## Where the model came from

The main article publishes neither equations nor parameter values. Everything in
this file is transcribed from the Supporting Information Heta export
(`PSP4-13-1798-s001.xlsx`) with the supplementary equation list
(`PSP4-13-1798-s002.docx`, S1-S28) used as an independent cross-check on
stoichiometry. **Where the executable export and the supplementary prose
disagree, the export is treated as authoritative**, because the paper directs the
reader to "the Model code" for the equations. Every such disagreement is listed
below.

## Differences between the published ODE list and the shipped model code

Four terms appear in the supplement's equation list but have no rate law
anywhere in the Heta export, so they are absent from this implementation:

1. **`V_el_aSyn_CSF`** (equation S1) - elimination of monomeric aSyn from CSF.
   No reaction, no rate constant. The oligomer analogue exists but is multiplied
   by `(0 + coef_PFFstr)`, so it too is zero except in the 48 h after a PFF
   injection. The consequence is that the untreated model has no bulk clearance
   of aSyn out of the CNS at all; extracellular pools equilibrate with cells
   rather than being drained. This is very likely why the simulated CSF monomer
   and ISF oligomer sit about 3-4x above the `asyn_M` (0.0011 nM) and `asyn_O`
   (0.00232 nM) reference constants. A plausible rate law would be
   `Q_CSF_PL * aSyn_CSF`, but it is not published and has **not** been added.
2. **`V_sec_aSyn_STR_BIF`** and **`V_sec_aSynO_STR_BIF`** (equations S2, S4, S12,
   S13) - secretion of monomeric and oligomeric aSyn from striatal neurons into
   *non-striatal* interstitial fluid. The Heta `Record` sheet defines their rate
   constants (`k_sec_aSyn_STR_BIF`, `k_sec_aSynO_STR_BIF`) but no reaction
   consumes them, so they are dead entries; only the fibrillar route
   (`V_sec_aSynFm_STR_BIF`) is implemented.
3. **`IM_CSF`** (equation S23) and its elimination `V_el_IM_CSF` - neither the
   species nor the rate law exists in the export. `IM_CSF` is a terminal sink
   that no other equation reads, so omitting it changes no model output; its
   elimination constant is unpublished in any case. The model therefore has 27
   states where the paper counts 28.

The one surviving member of that family, `k_sec_aSynFm_STR_BIF`, is **not** a
`Const` entry: `Record!k_sec_aSynFm_STR_BIF` is assigned `k_sec_aSynFm_BC_BIF`,
which resolves through `k_deg_aSynFm_BC / 3` to `k_deg_aSynO_BC / 10 / 3` =
0.00833 1/h. It is therefore derived in `model()` rather than given a numeric
value in `ini()`, so the chain stays visible and cannot drift from its source.

## Parameter values that differ between the prose and the code

| Quantity | Supplementary prose | Heta `Const`/`Record` (used here) |
|---|---|---|
| `k_diss_aSynFm` | "`k_diss_aSynO/10`" in one section, "`= k_diss_aSynO`" in another | `k_diss_aSynO / 100` |
| `k_deg_aSynFm_BC`, `k_deg_aSynFm_MG` | "equal to aSyn oligomers degradation" | `k_deg_aSynO_* / 10` |
| `k0_IM` | `Q_BIF_CSF / BIF / 1.7` | `Q_BIF_CSF * k_el_IM / BIF` |
| `BC_STR_coef` | "synthesis in striatum neurons is 5 times greater" | `4` (agrees with the main text's "fourfold") |
| IM secretion | prose adds a self-activation term `Emax_IM * IM / EC50_IM_MG` and no Hill exponents | Hill exponents `Nh_aSynFm_IM = 3`, `Nh_aSynO_IM = 2`, no self-activation term |
| Neurodegeneration hazard | linear sum of three drivers | full Emax / Hill ratio form |
| `V_olig_*`, `V_gr_*` | no antibody `Free_*` factors | `Free_aSynM_BC`, `Free_aSynO_BC` factors present (the prose describes the untreated model) |

The `k0_IM` row is the load-bearing one, and the baseline check above decides it:
the Const-sheet form yields a healthy-brain IM of
`r signif(f$IM_BIF, 4)` against the supplement's explicitly stated value of 1,
whereas the prose formula would give roughly 59.

## Features present in the export but not exercisable

* **PFF maturation downregulation is not implemented.** The paper says "we
  downregulated the maturation of aSyn PFFs during the first 2 weeks" and the
  export contains a `coef_mat_PFF` `TimeSwitcher` at t = 2336 h (exactly
  `time_PFFstr + 336 h` = 2 weeks), but it is marked inactive and **carries no
  assignment in any sheet**. The multiplier is therefore unpublished and has not
  been invented. This is the most likely reason the simulated post-PFF pathology
  decays back toward wild-type rather than showing the delayed rise in insoluble
  aSyn described for the 120-day time course (Figure S2e).
* **Antibody-specific affinities for prasinezumab and cinpanemab are absent.**
  The export contains only `Kd_mon` and `Kd_MEDI`, both 8 nM. The paper states
  that "prasinezumab and cinpanemab have higher affinities to aggregated aSyn"
  but the values are not in any on-disk source, so only the MEDI1341
  parameterisation is packaged. Simulating the other two antibodies requires
  overriding `Kd_MEDI` with values the paper does not publish.
* **The 2.6-fold late-phase viral overexpression is not switchable.** Table 1
  describes "threefold overexpression in the striatum during the first 3 weeks
  and 2.6-fold overexpression after 3 weeks"; only the 3-fold constant (`AAV`)
  exists, with no second switcher.
* **Four `Const` entries are unused by the shipped code**:
  `k_neurodeg_aSynO` (1.7e-6, superseded by the Emax form of the hazard, which
  uses `Emax_aSynO_neurodeg`), `aSyn_O` (1e-4, a stale duplicate of the
  `asyn_O` = 0.00232 nM actually referenced), and `time_BP` / `BP_period`, which
  no formula in any sheet references. All are omitted from `ini()` rather than
  carried as dead parameters.
* **Six further `Const` entries feed report-only normalisers, not dynamics**:
  `max_conc_O_BC`, `max_concO_MG`, `aSynF_1h`, `radio_aSyn_dose`, `PFF_ug_mL`
  and `aSyn_mon_ug_mL` appear only as denominators of percent-of-maximum
  `Record` expressions used for the in-vitro and radiolabel calibration figures.
  Those `Record`s are not part of the ODE system and are not shipped, so the
  constants are omitted too. Verified by searching every formula column of all
  seven sheets for each identifier.

## Reporting normalisers

`sol_aSyn_BL` (98) does not reproduce as a percentage: `perc_sol_aSyn_BR`
evaluates to about `r signif(f$perc_sol_aSyn_BR, 3)` at the wild-type baseline
rather than ~100%, so its scale is not consistent with the amount-valued
numerator of the `Record` expression. `insol_aSyn_BL` (1.895e-6) behaves as an
amount in nmol and gives `r sprintf("%.0f%%", f$perc_insol_aSyn_BR)` at baseline.
Both records are implemented verbatim, but relative-change validation in this
vignette uses fold-changes of the underlying concentration records rather than
these two normalisers.

## Encoding decisions

* **`SW_*` scenario switches are an encoding device, not paper parameters.** The
  Heta `TimeSwitcher` mechanism applies a one-off multiplicative assignment at a
  fixed time; rxode2 has no equivalent, so each is written as an algebraic step
  function gated by an `SW_*` flag. The switch times and multipliers are all
  paper values; only the on/off flags are new.
* **Antibody dosing is an event-table dose, not an assignment.** The export
  adds `Dose_Ig` to `Ig_PL` on a 168 h `TimeSwitcher` period; here that is a
  normal `ii = 168` IV dose record into `central`. Likewise the PFF challenge is
  a dose of `Dose_PFFstr` nM into the `aSynO_BIFstr` state.
* **`depot` is retained although it is inert.** Equation S24 carries an
  administration compartment with `V_abs_Ig = Ig_in` (an implicit 1/h absorption
  rate constant), but the export doses plasma directly, so `depot` stays at zero
  unless a user chooses to dose it.
* **Compartment naming.** The 20 aSyn pools, the two IM states and the hazard
  accumulator keep the paper's own symbols and are declared in
  `paper_specific_compartments`; this keeps the `model()` block diffable
  line-by-line against equations S1-S28. The four antibody states map onto
  canonical nlmixr2lib roles (`depot`, `central`, `peripheral1`, `isf`).
* **No IIV and no residual error.** The paper reports a single calibrated
  parameterisation with no variance components of any kind, so none are invented;
  every `ini()` entry is `fixed()`. The model is intended for deterministic
  simulation and cannot be fit as packaged.
* **The `aSynM_BC_BL`, `aSynO_BC_BL`, `asyn_M` and `asyn_O` constants are
  binding references, not baseline predictions.** The export's antibody-binding
  equations bind against these *frozen* constants rather than the dynamic
  states, so they are reference levels chosen by the authors and are not
  expected to equal the simulated wild-type values. They are not used as
  validation gates here.

## Quantitative agreement summary

| Published claim | Simulated | Verdict |
|---|---|---|
| Healthy-brain IM level = 1 | `r signif(f$IM_BIF, 4)` | agrees |
| Striatal aSyn fourfold above other regions (wt) | `r sprintf("%.2f", f$aSyn_STR / f$aSyn_BC)`-fold | agrees |
| Total brain aSyn just under 0.2 nM (wt) | `r sprintf("%.3f", f$aSyn_tot_BR)` nM | agrees |
| Total aSyn ~5x higher in transgenic mice | `r sprintf("%.2f", overexp$"Fold vs wild type"[2])`-fold | agrees |
| Oligomer ~10% (wt) to ~20% (tg) of total | `r sprintf("%.1f%%", f$aSynO_BR_perc)` to `r sprintf("%.1f%%", overexp$"Oligomer (%)"[2])` | broadly agrees |
| TLR2 knockout lowers aggregated aSyn 23-45% | `r sprintf("%.0f%% striatum, %.0f%% whole brain", -ko_tbl$"Aggregated aSyn, striatum (%)"[2], -ko_tbl$"Aggregated aSyn, brain (%)"[2])` (PFF arms) | agrees for PFF arms; transgenic-only arm shows ~1% |
| Inflammation inhibition alone: 25-30% reduction | `r sprintf("%.0f%%", -ai_tbl$"Aggregated aSyn, 30 dpi (%)"[2])` at 90% inhibition, started at the insult | agrees |
| Benefit only if started no later than the day of PFF | ~1% when started 1 week late | agrees |
| MoA1 alone increases aggregated aSyn | `r sprintf("%+.1f%%", moa_tbl$"Aggregated aSyn, 90 dpi (%)"[1])` | agrees |
| MoA2 near-100% reduction in oligomers | `r sprintf("%+.1f%%", moa_tbl$"Oligomer, 90 dpi (%)"[2])` | agrees |
| MoA3 ca. 60% reduction in aggregated aSyn | `r sprintf("%+.1f%%", moa_tbl$"Aggregated aSyn, 90 dpi (%)"[3])` | under-predicts |
| MoA4 near-100% reduction | `r sprintf("%+.1f%%", moa_tbl$"Aggregated aSyn, 90 dpi (%)"[4])` | agrees |
| MoA2 outcome dose-independent 10-50 mg/kg | dose-normalised AUC constant to `r sprintf("%.2g%%", 100 * max(abs(dn / dn[1] - 1)))` | consistent |
| Target engagement up to 80% for MEDI1341 | `r sprintf("%.0f%%", max(100 * (1 - te$Free_asyn[win])))` peak in mouse brain ISF | not comparable: the published figure is rat CSF |

The two shortfalls (MoA3, and the transgenic-only TLR2 arm) share a cause: the
published `EC50_IM_aggr` of 74.6 places the inflammation-to-aggregation coupling
well above the IM levels this parameterisation reaches unless a fibril challenge
is present. No parameter has been tuned to close either gap.
