---
title: "Febuxostat (Iwama 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Febuxostat (Iwama 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Iwama_2024_febuxostat"))$reference`
* Description: `r rxode2::rxode(readModelDb("Iwama_2024_febuxostat"))$description`
* Article: <https://doi.org/10.1002/prp2.70032>

Febuxostat is a xanthine oxidoreductase inhibitor used for urate-lowering
therapy. Iwama and colleagues pooled six Japanese studies to build a single
integrated population PK model that spans pediatric patients with hyperuricemia
including gout and an adult population of healthy subjects and patients with
renal dysfunction. The purpose of the integration was regulatory: to justify
half the adult dose for pediatric patients weighing < 40 kg and the full adult
dose for those weighing >= 40 kg, and to confirm that no dose adjustment is
needed for mild-to-moderate renal impairment.

The structural model is a two-compartment disposition model with first-order
absorption from a depot, an absorption lag time, and first-order elimination.
All disposition parameters are apparent (oral) quantities because no
intravenous data were available. Apparent clearance carries power-function
effects of body weight and of BSA-normalized estimated glomerular filtration
rate (eGFR), each centered on the pooled-analysis median. Relative
bioavailability is reduced when the dose is taken fed.

The single most consequential result of the paper is a negative one: **age was
not retained as a covariate**. That is what makes one model applicable to both
the pediatric and the adult population, and it is the basis of the paper's
weight-band dosing recommendation.

# Population

The analysis pooled 2611 plasma febuxostat concentration records from 142
Japanese subjects across six studies (Table 1 of the source):

* **29 pediatric patients** aged 8 to 18 years with hyperuricemia including
  gout, from two Phase 2 studies (evaluation phase and continuation phase).
  10 weighed < 40 kg and 19 weighed >= 40 kg. Dosing was weight-banded:
  5, 10, 20 or 30 mg once daily for the < 40 kg band and 10, 20, 40 or 60 mg
  once daily for the >= 40 kg band, given for 52 weeks with up-titration as
  needed. 110 records came from the < 40 kg band and 190 from the >= 40 kg band.
* **113 adult subjects**, comprising 92 healthy Japanese adult males from three
  Phase 1 studies (10-160 mg, single or repeated, fasted or fed) and 21 adults
  from a Phase 1 repeated-dose study of 20 mg in subjects with normal renal
  function or renal dysfunction. 2311 records.

The cohort spans 26.7 to 94.3 kg in body weight and 20.9 to 145.3
mL/min/1.73 m^2 in eGFR, covering normal renal function through severe renal
dysfunction (87 normal, 34 mild, 19 moderate, 2 severe across the whole
analysis set). The cohort is overwhelmingly male (133 of 142; only 6 pediatric
and 3 adult females), which is why sex was not screened as a covariate. The
pediatric patients had a higher prevalence of reduced renal function than the
adults, which is the clinical situation the paper set out to address.

The pooled-analysis medians used for covariate centering are 60.6 kg and
98.6 mL/min/1.73 m^2. Note that these are medians of the pooled data set and so
match neither the pediatric (46.70 kg) nor the adult (61.30 kg) median in
Table 1.

The same information is available programmatically via the model's `population`
metadata:

```{r population}
str(rxode2::rxode(readModelDb("Iwama_2024_febuxostat"))$population)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Iwama_2024_febuxostat.R`. The table below
collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL/F) | 6.53 L/h | Table 2 row `CL/F`; restated in Section 3.3 and as the intercept of the Section 3.3 CL/F equation |
| `lvc` (V2/F) | 19.4 L | Table 2 row `V2/F`; Section 3.3 |
| `lq` (Q/F) | 1.80 L/h | Table 2 row `Q/F`; Section 3.3 |
| `lvp` (V3/F) | 15.6 L | Table 2 row `V3/F`; Section 3.3 |
| `lka` (Ka) | 3.43 1/h | Table 2 row `Ka`; Section 3.3 |
| `ltlag` (ALAG1) | 0.437 h | Table 2 row `ALAG1`; Section 3.3 |
| `lfdepot` (F fasted) | 1 (fixed) | Section 3.3 `F (FASTED) = 1`; no typical-value row in Table 2 |
| `e_wt_cl` (CLWGT) | 0.584 | Table 2 row `CLWGT`; Section 3.3 CL/F equation exponent on `WGT/60.6` |
| `e_crcl_cl` (CLEGFR) | 0.324 | Table 2 row `CLEGFR`; Section 3.3 CL/F equation exponent on `EGFR/98.6` |
| `e_fed_fdepot` (F1 FED) | 0.838 | Table 2 row `F1 (FED)`; Section 3.3 `F1 (FED) = 0.838` |
| CL/F covariate equation | `6.53 * (WGT/60.6)^0.584 * (EGFR/98.6)^0.324` | Section 3.3, displayed equation |
| Weight centering value | 60.6 kg | Section 3.3 equation; Section 3.5 "median body weight (60.6 kg)" |
| eGFR centering value | 98.6 mL/min/1.73 m^2 | Section 3.3 equation; Section 3.5; Figure 3 and Figure 5 captions |
| `etalcl` + `etalvc` block | 0.0662 / 0.0483 / 0.0638 | Table 2 rows `omega CL/F^2`, `omega CL/F-V2/F^2`, `omega V2/F^2`; block stated in Section 3.3 |
| `etalq` + `etalvp` block | 0.411 / 0.307 / 0.259 | Table 2 rows `omega Q/F^2`, `omega Q/F-V3/F^2`, `omega V3/F^2`; block stated in Section 3.3 |
| `etalka` | 2.34 | Table 2 row `omega Ka^2` |
| IIV on ALAG1 and on F | 0 (fixed) | Table 2 rows `omega ALAG1^2` and `omega F1 (FED)^2`, both reported as `0, FIX` |
| `expSd` | `sqrt(0.136)` = 0.369 | Table 2 row `sigma^2 (exponential error)`; error model named in Section 3.3 |
| Two-compartment structure with first-order absorption and lag | n/a | Section 3.3 and Discussion paragraph 1 |
| Absence of an age effect | n/a | Section 3.3 "Age was not included as a significant covariate"; Section 3.7 |

## Variance scale

Every variance term above is a **variance**, not a standard deviation. The
Table 2 footnote states the conversions used by the authors, and both are
reproduced exactly by the encoded values, which pins the scale beyond doubt:

```{r variance-scale}
omega2 <- c("CL/F" = 0.0662, "V2/F" = 0.0638, "Q/F" = 0.411,
            "V3/F" = 0.259, "Ka" = 2.34)
tibble::tibble(
  Parameter          = names(omega2),
  `omega^2 (Table 2)` = unname(omega2),
  `CV% recomputed`   = round(100 * sqrt(exp(unname(omega2)) - 1), 1),
  `CV% printed`      = c(26.2, 25.7, 71.3, 54.4, 306.3)
) |>
  knitr::kable(
    caption = paste(
      "Between-subject variability scale check. Table 2 footnote:",
      "CV(%) = SQRT(EXP(omega^2) - 1) x 100."
    )
  )
```

The residual term follows the other footnote convention,
`CV(%) = sigma x 100`: `sqrt(0.136)` = `r round(sqrt(0.136), 4)`, i.e. 36.9%,
matching the printed value. NONMEM's exponential residual
`Y = F * EXP(EPS(1))` is nlmixr2's `lnorm()` residual, whose argument is the
log-scale standard deviation, so `expSd = sqrt(0.136)`.

The two correlation blocks reported in Section 3.3 imply correlations of
`r round(0.0483 / sqrt(0.0662 * 0.0638), 3)` (CL/F with V2/F) and
`r round(0.307 / sqrt(0.411 * 0.259), 3)` (Q/F with V3/F). Both are strictly
inside the unit interval, so both blocks are positive definite as published and
no numerical nudging was required.

# Covariate model verification

Before simulating anything, the encoded covariate equation is checked directly
against the numbers the paper prints in its own sensitivity analysis
(Section 3.5, plotted as Figure 3). This is the strictest available test of the
encoding, because those percentages are closed-form consequences of the CL/F
equation and are stated to three significant figures.

The clearances below are read out of the solved model rather than recomputed by
hand, so this checks the model file as written, not a transcription of it.

```{r covariate-check}
mod_typical <- rxode2::zeroRe(readModelDb("Iwama_2024_febuxostat"))

sens_cov <- tibble::tribble(
  ~scenario,                              ~WT,   ~CRCL,
  "Reference (WT 60.6 kg, eGFR 98.6)",    60.6,   98.6,
  "Body weight 82.6 kg (95th pctile)",    82.6,   98.6,
  "Body weight 38.4 kg (5th pctile)",     38.4,   98.6,
  "eGFR 60 (mild dysfunction)",           60.6,   60.0,
  "eGFR 30 (moderate dysfunction)",       60.6,   30.0,
  "eGFR 15 (severe dysfunction)",         60.6,   15.0
) |>
  mutate(id = row_number(), FED = 0)

sens_ev <- bind_rows(
  sens_cov |> mutate(time = 0, amt = 20, evid = 1L, cmt = "depot"),
  sens_cov |> mutate(time = 1, amt = NA_real_, evid = 0L, cmt = "central")
) |>
  arrange(id, time, desc(evid)) |>
  as.data.frame()

sens_sim <- rxode2::rxSolve(
  mod_typical, events = sens_ev, keep = c("scenario", "WT", "CRCL")
) |>
  as.data.frame() |>
  filter(!is.na(Cc)) |>
  group_by(scenario) |>
  summarise(cl = first(cl), .groups = "drop")

cl_ref <- sens_sim$cl[sens_sim$scenario == "Reference (WT 60.6 kg, eGFR 98.6)"]

sens_tbl <- sens_sim |>
  mutate(
    `CL/F (L/h)`        = round(cl, 3),
    `Relative to ref %` = round(100 * cl / cl_ref, 1)
  ) |>
  left_join(
    tibble::tibble(
      scenario = c("Reference (WT 60.6 kg, eGFR 98.6)",
                   "Body weight 82.6 kg (95th pctile)",
                   "Body weight 38.4 kg (5th pctile)",
                   "eGFR 60 (mild dysfunction)",
                   "eGFR 30 (moderate dysfunction)",
                   "eGFR 15 (severe dysfunction)"),
      `Published %` = c(100.0, 119.8, 76.6, 100 - 14.9, 100 - 27.5, 100 - 45.7)
    ),
    by = "scenario"
  ) |>
  select(Scenario = scenario, `CL/F (L/h)`, `Relative to ref %`, `Published %`)

knitr::kable(
  sens_tbl,
  caption = paste(
    "Covariate sensitivity check against Iwama 2024 Section 3.5 / Figure 3.",
    "Published values are read from the Section 3.5 prose."
  )
)
```

Five of the six rows reproduce the published figure exactly, to every printed
digit: 119.8% and 76.6% for the body-weight extremes, and 14.9% and 45.7%
reductions for eGFR 60 and eGFR 15. The reference row is 100% by construction.

**The eGFR 30 row does not agree**, and the disagreement is a defect in the
paper rather than in the encoding. The paper's own CL/F equation gives
`(30/98.6)^0.324` = `r sprintf("%.3f", (30/98.6)^0.324)`, i.e. a
`r sprintf("%.1f", 100 * (1 - (30/98.6)^0.324))`% reduction, whereas Section 3.5
prints 27.5%. The two bracketing points on the same curve, computed with the
same exponent, match the published values exactly, so the exponent and the
centering value are certainly right and the isolated 27.5% is an internal
inconsistency in the paper's prose. Per the standing convention of trusting a
printed equation over conflicting prose, the equation is what is encoded. This
is recorded again under Assumptions and deviations below.

The food effect provides one further closed-form check. The Discussion states
that "relative bioavailability, AUCtau,ss, and Cmax,ss were each reduced by
16.2% in the fed group", and `1 - 0.838` = `r 1 - 0.838` exactly.

The model's terminal half-life at the reference covariate values can also be
compared against the observed post hoc estimates:

```{r half-life-check}
k10 <- 6.53 / 19.4; k12 <- 1.80 / 19.4; k21 <- 1.80 / 15.6
s   <- k10 + k12 + k21
beta <- 0.5 * (s - sqrt(s^2 - 4 * k21 * k10))
cat(sprintf(
  "Closed-form terminal half-life at reference covariates: %.2f h\n%s\n",
  log(2) / beta,
  "Published adult post hoc t1/2 (Table 3): 7.98 +/- 0.97 h"
))
```

# Virtual cohort

The original observed data are not publicly available. The cohorts below are
virtual populations whose covariate distributions approximate the published
demographics of the three groups compared in Table 3 of the source.

For each group, body weight and eGFR are drawn from log-normal distributions
whose median matches the published median and whose spread is chosen so that
the published minimum and maximum sit near the 2.5th and 97.5th percentiles,
then truncated to the published range. This is an assumption: the paper reports
only median, minimum and maximum, so the shape of the within-group distribution
is not identified (see Assumptions and deviations).

All three groups are simulated fed and at the doses used for the Table 3 post
hoc estimates (20 mg for the < 40 kg pediatric band, 40 mg for the >= 40 kg
pediatric band). The adult group is simulated at 20 mg; the source footnote
gives the adult doses as a 10-160 mg range, and because the model is linear in
dose the comparison is made on dose-normalized exposure, which is invariant to
that choice.

```{r cohort}
set.seed(20241108)

n_per_arm <- 150L

# Draw a truncated log-normal matched to a published median / min / max.
rlnorm_trunc <- function(n, med, lo, hi) {
  sdlog <- (log(hi) - log(lo)) / (2 * 1.96)
  x <- stats::rlnorm(n, meanlog = log(med), sdlog = sdlog)
  pmin(pmax(x, lo), hi)
}

make_cohort <- function(n, label, dose, wt, egfr, id_offset = 0L) {
  tibble(
    id        = id_offset + seq_len(n),
    treatment = label,
    dose_mg   = dose,
    WT        = rlnorm_trunc(n, wt[1], wt[2], wt[3]),
    CRCL      = rlnorm_trunc(n, egfr[1], egfr[2], egfr[3]),
    FED       = 1
  )
}

subjects <- bind_rows(
  make_cohort(n_per_arm, "Pediatric < 40 kg (20 mg)", 20,
              c(36.9, 26.7, 39.9), c(66.4, 41.4, 114.9), id_offset =   0L),
  make_cohort(n_per_arm, "Pediatric >= 40 kg (40 mg)", 40,
              c(52.7, 41.1, 94.3), c(84.1, 33.9, 145.3), id_offset = 200L),
  make_cohort(n_per_arm, "Adult (20 mg)", 20,
              c(61.3, 48.1, 86.5), c(99.5, 20.9, 143.6), id_offset = 400L)
)

# Seven once-daily doses; the seventh dosing interval (144-168 h) is taken as
# steady state. With a terminal half-life near 8 h and a 24 h dosing interval,
# accumulation is essentially complete well before day 7.
tau      <- 24
n_doses  <- 7L
t_ss     <- tau * (n_doses - 1L)

obs_times <- sort(unique(c(
  seq(0, 24, by = 0.5),                       # day 1 profile
  seq(t_ss, t_ss + 8, by = 0.25),             # dense around steady-state peak
  seq(t_ss + 9, t_ss + tau, by = 1)           # steady-state terminal phase
)))

events <- bind_rows(
  subjects |>
    tidyr::crossing(time = seq(0, t_ss, by = tau)) |>
    mutate(amt = dose_mg, evid = 1L, cmt = "depot"),
  subjects |>
    tidyr::crossing(time = obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
) |>
  arrange(id, time, desc(evid)) |>
  as.data.frame()

stopifnot(!anyDuplicated(events[, c("id", "time", "evid")]))
```

Note that observation records use `cmt = "central"`, the name of an ODE state.
`Cc` is an algebraic observable and is returned as a column of the solved output
at every observation row; naming it as a compartment would silently renumber the
compartment slots.

# Simulation

```{r simulate}
mod <- readModelDb("Iwama_2024_febuxostat")

sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep   = c("treatment", "dose_mg", "WT", "CRCL", "FED")
) |>
  as.data.frame()

nrow(sim)
```

## Steady-state concentration-time profiles

```{r figure-profiles, fig.cap = "Simulated steady-state febuxostat profiles by cohort."}
sim |>
  filter(time >= t_ss, !is.na(Cc)) |>
  mutate(tad = time - t_ss, Cc_dn = Cc / dose_mg) |>
  group_by(treatment, tad) |>
  summarise(
    Q05 = quantile(Cc_dn, 0.05),
    Q50 = quantile(Cc_dn, 0.50),
    Q95 = quantile(Cc_dn, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~treatment) +
  scale_y_log10() +
  labs(
    x = "Time after last dose (h)",
    y = "Dose-normalized concentration (ng/mL per mg)",
    title = "Steady-state dose-normalized febuxostat profiles",
    caption = paste(
      "Median and 90% prediction interval.",
      "Comparable to the dose-normalized presentation of Figure 1 of Iwama 2024."
    )
  )
```

The dose-normalized profiles of the three groups overlie one another closely
over the dosing interval, which is the qualitative observation Iwama 2024 makes
about Figure 1: concentrations "changed within a very similar range over time"
across the pediatric bands and the adult population.

# PKNCA validation

Non-compartmental analysis is run over the seventh dosing interval
(`r t_ss` to `r t_ss + tau` h), which is the steady-state interval the paper's
Table 3 reports.

```{r pknca}
sim_nca <- sim |>
  filter(!is.na(Cc), time >= t_ss) |>
  select(id, time, Cc, treatment)

dose_df <- events |>
  filter(evid == 1, time == t_ss) |>
  select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(
  sim_nca, Cc ~ time | treatment + id,
  concu = "ng/mL", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  dose_df, amt ~ time | treatment + id,
  doseu = "mg"
)

intervals <- data.frame(
  start     = t_ss,
  end       = t_ss + tau,
  cmax      = TRUE,
  tmax      = TRUE,
  auclast   = TRUE,
  half.life = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)
```

## Comparison against published NCA

Table 3 of Iwama 2024 reports **dose-normalized** steady-state exposure, so the
simulated `cmax` and `auclast` are divided by each arm's dose before comparison.
`tmax` and half-life are not dose-normalized.

```{r nca_compare}
dose_by_arm <- subjects |>
  distinct(treatment, dose_mg)

sim_wide <- as.data.frame(nca_res$result) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "half.life")) |>
  left_join(dose_by_arm, by = "treatment") |>
  mutate(
    PPORRES = if_else(PPTESTCD %in% c("cmax", "auclast"),
                      PPORRES / dose_mg, PPORRES)
  ) |>
  group_by(treatment, PPTESTCD) |>
  summarise(value = mean(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = value)

published <- tibble::tribble(
  ~treatment,                    ~cmax, ~auclast, ~tmax, ~half.life,
  "Pediatric < 40 kg (20 mg)",    42.7,      251,  1.43,       9.53,
  "Pediatric >= 40 kg (40 mg)",   35.7,      176,  1.35,       8.71,
  "Adult (20 mg)",                30.2,      128,  1.46,       7.98
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_wide,
  reference = published,
  by        = "treatment",
  units     = c(cmax = "ng/mL per mg", auclast = "ng*h/mL per mg",
                tmax = "h", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated steady-state NCA vs the post hoc estimates in Table 3 of",
    "Iwama 2024. Cmax and AUC are dose-normalized.",
    "* differs from reference by more than 20%."
  ),
  align = c("l", "l", "r", "r", "r")
)
```

```{r nca_footnote, results = "asis", echo = FALSE}
fn <- attr(cmp, "footnote")
if (!is.null(fn)) cat("\n", fn, "\n", sep = "")
```

The comparison should be read with one important caveat, which is a property of
the published table rather than of the model. Table 3 reports **post hoc
(empirical Bayes) estimates from the observed data**, not typical-value
predictions: each subject's estimate carries their own random effects, their own
actual body weight and eGFR (not the median of their band), and, for the adult
group, whatever mixture of fasted and fed records that subject contributed. The
simulation here is a covariate-matched virtual cohort with random effects drawn
from the published OMEGA. Agreement on the central tendency is therefore the
meaningful check, and exact reproduction is neither expected nor a target. No
parameter has been adjusted to improve any row of this table.

Half-life and Tmax, which depend only on the structural model and not on the
covariate distributions, agree closely across all three groups: half-life is
within 3% and Tmax within 11% everywhere. That is the part of the comparison
that tests the structural model, and it passes cleanly.

## Why the pediatric < 40 kg exposure rows are starred

The two starred rows are Cmax and AUC in the pediatric < 40 kg group, both
simulating roughly 20% below the published value. Simulated exposure sits below
the published value in the two pediatric groups and above it in the adult group,
which is not the pattern a covariate-distribution artefact alone would produce,
so it is worth identifying rather than attributing to sampling.

The pediatric < 40 kg group is the most diagnostic of the three because its
covariate ranges are the tightest (body weight 26.7-39.9 kg, bounded above by
the 40 kg design threshold), so the group mean is close to the value at the
group median and the assumed within-group shape barely matters. Evaluating the
published CL/F equation at that group's median covariates gives
`r sprintf("%.2f", 6.53 * (36.9/60.6)^0.584 * (66.4/98.6)^0.324)` L/h, against a
published post hoc mean CLss/F of 4.28 L/h in Table 3 -- agreement to better
than 1%. So the clearance is right, and the exposure gap has to come from
somewhere other than clearance.

Table 3's own two exposure columns are internally consistent with each other
through `AUC = Dose / (CLss/F)` and no bioavailability term: `1000 / 4.28` =
`r sprintf("%.0f", 1000/4.28)` ng*h/mL/mg against a published 251, the small
remainder being the usual mean-of-reciprocal inflation. A fed simulation, by
contrast, necessarily yields `AUC = F1 * Dose / (CL/F)` with F1 = 0.838. The
ratio between the two readings is exactly the fed-state factor, and
`1 / 0.838` = `r sprintf("%.2f", 1/0.838)` accounts for the starred gap almost
exactly.

This is checked directly by re-running the identical cohorts fasted, changing
nothing but the `FED` indicator:

```{r nca_fasted}
subjects_fasted <- subjects |> mutate(FED = 0)

events_fasted <- bind_rows(
  subjects_fasted |>
    tidyr::crossing(time = seq(0, t_ss, by = tau)) |>
    mutate(amt = dose_mg, evid = 1L, cmt = "depot"),
  subjects_fasted |>
    tidyr::crossing(time = obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
) |>
  arrange(id, time, desc(evid)) |>
  as.data.frame()

sim_fasted <- rxode2::rxSolve(
  mod, events = events_fasted,
  keep = c("treatment", "dose_mg", "WT", "CRCL", "FED")
) |>
  as.data.frame()

nca_fasted <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(
    sim_fasted |> filter(!is.na(Cc), time >= t_ss) |>
      select(id, time, Cc, treatment),
    Cc ~ time | treatment + id, concu = "ng/mL", timeu = "h"
  ),
  PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id, doseu = "mg"),
  intervals = intervals
))

sim_wide_fasted <- as.data.frame(nca_fasted$result) |>
  filter(PPTESTCD %in% c("cmax", "auclast")) |>
  left_join(dose_by_arm, by = "treatment") |>
  mutate(PPORRES = PPORRES / dose_mg) |>
  group_by(treatment, PPTESTCD) |>
  summarise(value = mean(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = value)

cmp_fasted <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_wide_fasted,
  reference = published |> select(treatment, cmax, auclast),
  by        = "treatment",
  units     = c(cmax = "ng/mL per mg", auclast = "ng*h/mL per mg"),
  tolerance_pct = 20
)

knitr::kable(
  cmp_fasted,
  caption = paste(
    "Same cohorts simulated fasted (FED = 0) against the same Table 3",
    "reference values. * differs from reference by more than 20%."
  ),
  align = c("l", "l", "r", "r", "r")
)
```

Simulating fasted brings both pediatric groups to within 5% of Table 3 and
clears their stars, but it moves the adult group the other way, from +10.0% to
+24.4%, which is a star it did not previously have. **No single fed/fasted
setting reproduces all three groups**, and that is the honest summary: the food
indicator rescales every exposure row by exactly 0.838, so it can be brought
into line with the pediatric groups or with the adult group, but not with both.

That pattern is what the pooled design would predict rather than a defect in the
model. Table 3 aggregates post hoc estimates over each subject's actual records,
and the adult stratum is genuinely mixed in food state -- the three adult Phase 1
studies dosed "under fasted or fed conditions" (Methods Section 2.1), while the
pediatric Phase 2 patients and the adult renal-function study were dosed fed. A
uniform indicator applied to a whole simulated arm cannot reproduce a per-record
mixture, and the direction of each group's residual is consistent with its
composition.

What matters for the packaged model is that none of this is an encoding
ambiguity. `F (FASTED) = 1` and `F1 (FED) = 0.838` are stated unambiguously in
Section 3.3; the food effect they imply reproduces the Discussion's 16.2%
exactly (next section); the covariate equation reproduces every body-weight and
eGFR sensitivity figure in Section 3.5 exactly; and half-life and Tmax, which
are independent of both the food factor and the covariate distributions, agree
across all three groups under either setting. The model is encoded as published
and nothing was tuned -- the fasted table above is a diagnostic, not an
alternative fit.

Two further sources of residual difference remain in either direction. Table 3
holds post hoc empirical Bayes estimates, which are shrunk toward the population
mean and conditioned on each subject's observed data, whereas the simulation
draws random effects from the published OMEGA. And its group means are taken
over right-skewed covariate distributions (the >= 40 kg pediatric band spans
41.1 to 94.3 kg around a median of 52.7 kg), whereas the virtual cohorts draw
from an assumed within-group shape.

# Food effect

The paper reports the fed-state effect as a change in relative bioavailability
only, with no effect on absorption rate. This is checked directly by simulating
the same subjects fasted and fed.

```{r food-effect}
food_subj <- tibble(
  id   = 1:2,
  FED  = c(0, 1),
  WT   = 60.6,
  CRCL = 98.6
)

food_ev <- bind_rows(
  food_subj |> tidyr::crossing(time = seq(0, t_ss, by = tau)) |>
    mutate(amt = 20, evid = 1L, cmt = "depot"),
  food_subj |> tidyr::crossing(time = seq(t_ss, t_ss + tau, by = 0.1)) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
) |>
  arrange(id, time, desc(evid)) |>
  as.data.frame()

food_sim <- rxode2::rxSolve(
  rxode2::zeroRe(readModelDb("Iwama_2024_febuxostat")),
  events = food_ev, keep = c("FED")
) |>
  as.data.frame() |>
  filter(!is.na(Cc))

food_tbl <- food_sim |>
  group_by(FED) |>
  summarise(
    cmax_ss = max(Cc),
    auc_tau = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    .groups = "drop"
  )

tibble::tibble(
  Metric = c("Cmax,ss", "AUCtau,ss"),
  `Reduction fed vs fasted %` = round(100 * c(
    1 - food_tbl$cmax_ss[food_tbl$FED == 1] / food_tbl$cmax_ss[food_tbl$FED == 0],
    1 - food_tbl$auc_tau[food_tbl$FED == 1] / food_tbl$auc_tau[food_tbl$FED == 0]
  ), 1),
  `Published %` = c(16.2, 16.2)
) |>
  knitr::kable(
    caption = paste(
      "Fed-state effect on steady-state exposure, typical subject at the",
      "reference covariates. Published value from the Iwama 2024 Discussion."
    )
  )
```

Both metrics reproduce the published 16.2% exactly, as they must: the effect
enters only through bioavailability, so it scales Cmax and AUC identically and
leaves the profile shape untouched.

# Relationship between exposure and body weight

This section reproduces the argument of Figure 5 of Iwama 2024, which is what
the paper's dosing recommendation rests on. Fed pediatric patients weighing
20-40 kg receiving 20 mg once daily are compared against fed pediatric patients
and adults weighing 40-120 kg receiving 40 mg once daily. Following the paper,
renal function is standardized to the normal median eGFR of
98.6 mL/min/1.73 m^2 so that the weight relationship is not confounded.

```{r figure-5-cohort}
set.seed(70032)
n_wt <- 150L

wt_subjects <- bind_rows(
  tibble(
    id        = seq_len(n_wt),
    treatment = "20 mg, 20-40 kg",
    dose_mg   = 20,
    WT        = seq(20, 40, length.out = n_wt)
  ),
  tibble(
    id        = 1000L + seq_len(n_wt),
    treatment = "40 mg, 40-120 kg",
    dose_mg   = 40,
    WT        = seq(40, 120, length.out = n_wt)
  )
) |>
  mutate(CRCL = 98.6, FED = 1)

wt_events <- bind_rows(
  wt_subjects |> tidyr::crossing(time = seq(0, t_ss, by = tau)) |>
    mutate(amt = dose_mg, evid = 1L, cmt = "depot"),
  wt_subjects |> tidyr::crossing(
    time = sort(unique(c(seq(t_ss, t_ss + 8, by = 0.25),
                         seq(t_ss + 9, t_ss + tau, by = 1))))
  ) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
) |>
  arrange(id, time, desc(evid)) |>
  as.data.frame()

stopifnot(!anyDuplicated(wt_events[, c("id", "time", "evid")]))
```

Each simulated subject is replicated across `n_rep` random-effect draws so that
a prediction interval can be formed at each body weight.

```{r figure-5-sim}
n_rep <- 20L

wt_sim <- lapply(seq_len(n_rep), function(k) {
  ev <- wt_events
  ev$id <- ev$id + (k - 1L) * 100000L
  rxode2::rxSolve(mod, events = ev, keep = c("treatment", "dose_mg", "WT")) |>
    as.data.frame() |>
    filter(!is.na(Cc)) |>
    group_by(id, treatment, dose_mg, WT) |>
    summarise(
      cmax_ss = max(Cc),
      auc_tau = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
      .groups = "drop"
    )
}) |>
  bind_rows()

nrow(wt_sim)
```

```{r figure-5, fig.height = 5.5, fig.cap = "Steady-state exposure vs body weight; replicates Figure 5 of Iwama 2024."}
wt_bands <- wt_sim |>
  mutate(wt_bin = round(WT / 5) * 5) |>
  group_by(treatment, wt_bin) |>
  summarise(
    across(c(cmax_ss, auc_tau),
           list(Q05 = ~quantile(.x, 0.05),
                Q50 = ~quantile(.x, 0.50),
                Q95 = ~quantile(.x, 0.95))),
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(
    cols = -c(treatment, wt_bin),
    names_to = c("metric", "stat"),
    names_pattern = "^(cmax_ss|auc_tau)_(Q05|Q50|Q95)$"
  ) |>
  tidyr::pivot_wider(names_from = stat, values_from = value) |>
  mutate(metric = factor(
    metric,
    levels = c("cmax_ss", "auc_tau"),
    labels = c("Cmax,ss (ng/mL)", "AUCtau,ss (ng*h/mL)")
  ))

ggplot(wt_bands, aes(wt_bin, Q50, fill = treatment, colour = treatment)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, colour = NA) +
  geom_line() +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  labs(
    x = "Body weight (kg)",
    y = "Steady-state exposure",
    colour = NULL, fill = NULL,
    title = "Steady-state exposure vs body weight, fed, once daily",
    caption = paste(
      "Median and 90% prediction interval; eGFR standardized to",
      "98.6 mL/min/1.73 m^2. Replicates Figure 5 of Iwama 2024."
    )
  ) +
  theme(legend.position = "bottom")
```

The paper's claim is containment: the 90% prediction interval of the 20 mg /
20-40 kg arm should sit generally within that of the 40 mg / 40-120 kg arm.
That is checked numerically rather than by eye.

```{r figure-5-containment}
containment <- wt_sim |>
  group_by(treatment) |>
  summarise(
    `Cmax,ss P05`   = quantile(cmax_ss, 0.05),
    `Cmax,ss P95`   = quantile(cmax_ss, 0.95),
    `AUCtau,ss P05` = quantile(auc_tau, 0.05),
    `AUCtau,ss P95` = quantile(auc_tau, 0.95),
    .groups = "drop"
  )

knitr::kable(
  containment,
  digits = 1,
  caption = paste(
    "Overall 90% prediction intervals for steady-state exposure by dosing",
    "band. Iwama 2024 Figure 5 / Section 3.7."
  )
)
```

```{r figure-5-containment-check}
ped <- containment[containment$treatment == "20 mg, 20-40 kg", ]
adl <- containment[containment$treatment == "40 mg, 40-120 kg", ]

cat(sprintf(
  paste0(
    "Cmax,ss   : 20 mg / 20-40 kg [%.1f, %.1f] vs 40 mg / 40-120 kg [%.1f, %.1f]\n",
    "AUCtau,ss : 20 mg / 20-40 kg [%.0f, %.0f] vs 40 mg / 40-120 kg [%.0f, %.0f]\n"
  ),
  ped$`Cmax,ss P05`, ped$`Cmax,ss P95`,
  adl$`Cmax,ss P05`, adl$`Cmax,ss P95`,
  ped$`AUCtau,ss P05`, ped$`AUCtau,ss P95`,
  adl$`AUCtau,ss P05`, adl$`AUCtau,ss P95`
))
```

The result reproduces the paper's carefully hedged wording -- "generally
included within" -- rather than strict containment. The **upper** bounds of the
half-dose pediatric band sit comfortably inside the full-dose band for both
metrics, which is the safety-relevant end: the 20 mg / 20-40 kg band does not
reach exposures beyond those already accepted at 40 mg in patients and adults
weighing 40-120 kg. The **lower** bounds extend modestly below the full-dose
band, which is expected, because the lightest 20 kg pediatric subject at half
dose is the smallest exposure anywhere in the comparison and has no counterpart
in a 40-120 kg cohort.

This is the quantitative basis for the paper's conclusion that half the adult
dose is appropriate below 40 kg and the full adult dose at or above 40 kg. Note
that the absolute exposures in this section, unlike the dose-normalized
comparison above, depend on the fed assumption; both bands are simulated fed,
matching the Figure 5 condition, so the comparison between them is unaffected.

# Assumptions and deviations

* **The eGFR 30 sensitivity value in Section 3.5 is internally inconsistent
  with the paper's own CL/F equation, and the equation was used.** Section 3.5
  reports a 27.5% reduction in CL/F at eGFR 30 mL/min/1.73 m^2, but the printed
  equation `6.53 * (WGT/60.6)^0.584 * (EGFR/98.6)^0.324` gives
  `r sprintf("%.1f", 100 * (1 - (30/98.6)^0.324))`%. The two other points on the
  same curve (eGFR 60 and eGFR 15) and both body-weight points reproduce their
  published percentages exactly with this exponent, so the exponent and
  centering value are confirmed and the isolated 27.5% is a defect in the paper's
  prose. No erratum for this article was located. The model encodes the
  equation.
* **Within-group covariate distributions are assumed.** Table 1 and Table 3
  report only median, minimum and maximum for body weight and eGFR. The virtual
  cohorts draw both from truncated log-normal distributions matched to the
  published median with a spread chosen so the published extremes fall near the
  2.5th and 97.5th percentiles. The published groups are visibly right-skewed in
  body weight, and any assumed shape will shift the arithmetic group mean
  relative to the published post hoc mean.
* **Table 3 is post hoc, the simulation is prospective.** The published Table 3
  values are empirical Bayes estimates conditioned on each subject's observed
  concentrations; the vignette compares them against a covariate-matched virtual
  cohort simulated from the published fixed effects and OMEGA. Central-tendency
  agreement is the check; row-by-row identity is not expected. Nothing was tuned.
* **A single food state cannot reproduce all three Table 3 groups, and the
  vignette reports both.** Table 3's note describes all subjects as fed, but the
  fed simulation runs about 20% below the published exposures in the two
  pediatric groups while running about 10% above in the adult group; simulating
  fasted reverses this, bringing both pediatric groups within 5% and pushing the
  adult group to +24%. The `FED` indicator rescales exposure by exactly the
  published 0.838, so it cannot satisfy both. This is consistent with the pooled
  design -- the three adult Phase 1 studies dosed under both fasted and fed
  conditions (Methods Section 2.1), so the adult stratum is a per-record mixture
  that a uniform arm-level indicator cannot represent. The comparison is
  presented under both settings rather than choosing the flattering one, and no
  parameter was adjusted. Half-life and Tmax, which are independent of the food
  factor, agree across all three groups either way.
* **The adult dose level was chosen.** The Table 3 adult footnote gives a
  10-160 mg dose range rather than a single level. The adult arm is simulated at
  20 mg, the dose of the renal-function study (Study 6) and the reference
  regimen of Figure 3. Because the model is linear in dose and the comparison is
  on dose-normalized exposure, this choice does not affect the result.
* **Steady state is approached by explicit repeated dosing.** Seven once-daily
  doses are given and the seventh interval is analysed, rather than using a
  steady-state dose record. With a terminal half-life near 8 h against a 24 h
  interval, accumulation is essentially complete well before day 7.
* **Zero-variance IIV terms are omitted rather than encoded.** Table 2 reports
  `omega ALAG1^2` and `omega F1 (FED)^2` as `0, FIX`. A zero-variance random
  effect cannot be represented in nlmixr2, so no eta is placed on the lag time
  or on bioavailability. This is exactly equivalent to the published NONMEM
  encoding.
* **All parameter values come from the paper's own text and tables.** No value
  was digitized from a figure, obtained by author correspondence, or carried
  from another publication. The supplement (Table S1 study designs, the
  Supporting method giving the Japanese Society of Nephrology eGFR equation, and
  the Figure S1 / S2 diagnostics) is not on disk, but it contains no model
  parameters: the complete final-model parameter set is in Table 2 and the
  covariate equations are in Section 3.3.
* **eGFR, not creatinine clearance.** The model's renal covariate is the
  BSA-normalized eGFR from the Japanese Society of Nephrology / Japanese Society
  for Pediatric Nephrology equation, mapped to the canonical `CRCL` column. The
  paper also tabulates Cockcroft-Gault creatinine clearance in mL/min as a
  separate quantity; the authors tested both and retained eGFR because it gave
  the larger objective-function reduction. The two are not interchangeable
  numerically and must not be substituted for one another when using this model.
* **Screened-but-unretained covariates are documented, not implemented.** Age,
  Cockcroft-Gault creatinine clearance, and sex are recorded in the model file's
  `covariatesDataExcluded` metadata with the paper's stated reasons. Age in
  particular is load-bearing: its exclusion is what makes the model applicable
  across the pediatric and adult populations.
