---
title: "Aldehyde oxidase substrates: intravenous PK (Izat 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aldehyde oxidase substrates: intravenous PK (Izat 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Models and source

This paper contributes three models, one per compound:

```{r model-list}
model_names <- c("Izat_2025_zaleplon",
                 "Izat_2025_ziprasidone",
                 "Izat_2025_zoniporide")
uis <- lapply(model_names, function(n) rxode2::rxode(readModelDb(n)))
names(uis) <- model_names
```

* Citation: `r uis[["Izat_2025_zaleplon"]]$reference`
* Article: <https://doi.org/10.1002/psp4.13255>
* Supplement (Appendix S1): <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC11706420/>

# What these models are, and what they are not

Izat 2025 is a **framework paper**. It builds Simcyp (version 21) PBPK models
for six aldehyde-oxidase (AO) substrates -- capmatinib, idelalisib, lenvatinib,
zaleplon, ziprasidone and zoniporide -- in order to ask whether PBPK can
predict AO-mediated clearance, the fraction metabolised by AO (`fmAO`) versus
by CYP3A4 (`fmCYP3A4`), and the resulting CYP3A4 drug-drug-interaction risk.

Simcyp's whole-body mass-balance equations are proprietary and appear nowhere
in the paper or Appendix S1. The corpus contains sixteen numbered equations and
not one of them is a disposition equation: Equations S1-S7 and S10-S13 are
bench-assay arithmetic, Equation S8 is a permeability correlation, and
Equations S14-S16 are a *static* (time-free, algebraic) AO-DDI calculator. The
platform model therefore cannot be encoded as published.

What *is* completely reported, in Table 1, is each compound's **disposition
card**: the steady-state volume of distribution, the volume of the single
adjusting compartment (SAC) and the inter-compartmental clearance where a
minimal-PBPK layout was used, the observed intravenous plasma clearance, and
the observed renal clearance. Appendix S1 Supplement 2 states the minimal-PBPK
structure in the paper's own words -- "a non-physiological central compartment
excluding the liver and portal vein coupled with a non-physiological
non-central compartment, facilitating adjustment of systemic compartment
concentration profile (SAC)". That is enough to rebuild the **intravenous**
disposition of three of the six compounds as ordinary compartmental models.

## Why only the intravenous arms

The oral arms of this paper are deliberately **not** reduced. Oral
bioavailability in the source is `F = fa * Fg * fh`, and while Table 1 prints
`fa` and Appendix S1 gives `Fg`, the hepatic availability `fh = 1 - CL/Q_H`
requires the hepatic blood flow `Q_H`. `Q_H` is mentioned twice in the prose
and **never given a value**; neither is liver weight, and neither is body
weight. Supplying `Q_H` from the Simcyp platform default would be exactly the
substitution that this package's PBPK sourcing rule forbids -- and it is not a
rounding-level choice, because moving `Q_H` from 70 to 110 L/h moves the
predicted zaleplon oral Cmax from 0.62 to 1.29 times observed.

The intravenous arms have no such dependency. They need the reported clearance,
the reported volume, and one assumption -- a 70 kg reference body weight, since
Table 1 states every volume in L/kg and no weight is printed. That is the
standing "undefined reference value maps to a rounded standard" convention, and
it is the *only* external input anywhere in the three models below.

Capmatinib, idelalisib and lenvatinib are oral-only in this paper and are not
extracted from it. Capmatinib additionally fails a cross-dose consistency
check: its observed oral clearance is 36.1 L/h at 200 mg and 51.7 L/h at
600 mg, genuinely nonlinear, while Table 1 carries a single weighted-mean
42.3 L/h that cannot serve both doses.

## What is not reproducible

**Every headline result of the paper.** The `fmAO` / `fmCYP3A4` split, the
empirical scaling factors applied to in vitro intrinsic clearance, and all of
the CYP3A4 DDI area-ratio predictions against ketoconazole, itraconazole,
rifampicin and carbamazepine depend on the unpublished Simcyp enzyme-abundance
layer and on proprietary Certara perpetrator compound files. These models
reproduce the intravenous control profiles only. See
[Assumptions and deviations](#assumptions-and-deviations).

# Population

Each compound's intravenous data come from a single published study, listed in
Appendix S1 Table S3.

```{r population}
pop_tab <- do.call(rbind, lapply(model_names, function(n) {
  p <- uis[[n]]$population
  data.frame(
    Model      = n,
    Species    = p$species,
    N          = p$n_subjects,
    `Age (y)`  = p$age_range,
    `% female` = p$sex_female_pct,
    Dose       = p$dose_range,
    check.names = FALSE
  )
}))
knitr::kable(pop_tab, row.names = FALSE,
             caption = "Population per compound (Izat 2025 Appendix S1 Table S3).")
```

* **Zaleplon** -- Rosen 1999 (Appendix S1 Table S3 reference 22). Ten healthy
  volunteers aged 19-32 y received 5 mg intravenously. Table S3 records the
  female fraction as "0 or 1", i.e. separate all-female and all-male cohorts,
  which is why Table 1 reports two clearances for this compound and no other.
* **Ziprasidone** -- Miceli 2005 (reference 23). Twelve healthy men aged
  19-37 y received 5 mg intravenously.
* **Zoniporide** -- Dalvie 2010 (reference 13). Four healthy men aged 18-55 y
  received 80 mg intravenously in a combined pharmacokinetic and mass-balance
  study.

None of the three studies reports a body weight, and neither does Appendix S1.

# Source trace

Every value in every `ini()` block, with its source location.

```{r source-trace}
trace_tab <- tibble::tribble(
  ~Model,        ~Parameter,    ~Value,     ~Source,
  "zaleplon",    "lvc",         "81.1993 L", "Table 1: Vss 1.16 L/kg, Vsac 0.00001 L/kg; (1.16 - 0.00001) * 70 kg",
  "zaleplon",    "lcl",         "71.58 L/h", "Table 1 'CLiv/oral (L/h) [CV%]': 71.58 [20.2] (i.v., male)",
  "zaleplon",    "e_sexf_cl",   "log(0.733585)", "Table 1: 52.51 L/h (i.v., female) / 71.58 L/h (i.v., male)",
  "zaleplon",    "propSd",      "0 (fixed)", "No residual-error model is reported anywhere in the source",
  "ziprasidone", "lvc",         "70.7 L",    "Table 1: Vss 1.01 L/kg; 1.01 * 70 kg",
  "ziprasidone", "lcl",         "23.04 L/h", "Table 1 'CLiv/oral (L/h) [CV%]': 23.04 [14] (i.v.)",
  "ziprasidone", "propSd",      "0 (fixed)", "No residual-error model is reported anywhere in the source",
  "zoniporide",  "lvc",         "119.0 L",   "Table 1: Vss 1.70 L/kg; 1.70 * 70 kg",
  "zoniporide",  "lcl_renal",   "16.2 L/h",  "Table 1 'CLRenal (L/h)': 16.2",
  "zoniporide",  "lcl_nonren",  "79.15 L/h", "Table 1: CLiv 95.35 [5.7] minus CLRenal 16.2",
  "zoniporide",  "propSd",      "0 (fixed)", "No residual-error model is reported anywhere in the source"
)
knitr::kable(trace_tab, caption = "Source trace for every fixed parameter.")
```

The model equations are the standard one-compartment intravenous system
(`d/dt(central) = -kel * central`, `kel = cl / vc`, `Cc = 1000 * central / vc`)
with no nonstandard terms. The `1000` converts mg/L to the ng/mL used
throughout Appendix S1 Table S10. Each compound is one compartment for a
different, explicitly stated reason:

* **Zaleplon** is classified "Minimal PBPK" in Table 1, but its reported SAC
  parameters are `Vsac = 0.00001 L/kg` and `Q = 0 L/h`. An inter-compartmental
  clearance of exactly zero means the SAC never receives drug, so the reduction
  to one compartment is **exact**, not an approximation.
* **Ziprasidone** and **zoniporide** are classified "Full PBPK". In a full-PBPK
  layout the multi-tissue structure *is* the distribution model, and the only
  distribution quantity Table 1 reports is the lumped steady-state volume that
  the fitted global partition-coefficient scalar was tuned to recover. No
  tissue volumes, blood flows or partition coefficients are printed, and no
  `Vsac` or `Q` is given. These two reductions are therefore genuinely
  mono-exponential approximations of multi-exponential profiles, and the cost
  shows up in Cmax below (about 15% low and 14% high respectively). That is
  reported, not tuned away.

## Reproducing the derivations

```{r derivation}
BW <- 70  # kg -- assumed reference weight; see Errata

# Table 1 inputs, verbatim.
tab1 <- tibble::tribble(
  ~drug,          ~Vss_Lkg, ~Vsac_Lkg, ~CL_Lh,  ~CLrenal_Lh,
  "zaleplon_F",       1.16,   0.00001,  52.51,          0,
  "zaleplon_M",       1.16,   0.00001,  71.58,          0,
  "ziprasidone",      1.01,        NA,  23.04,          0,
  "zoniporide",       1.70,        NA,  95.35,       16.2
)

derived <- tab1 |>
  dplyr::mutate(
    vc_L      = (Vss_Lkg - dplyr::coalesce(Vsac_Lkg, 0)) * BW,
    kel_perh  = CL_Lh / vc_L,
    thalf_h   = log(2) / kel_perh
  )
knitr::kable(derived |> dplyr::mutate(dplyr::across(where(is.numeric),
                                                    \(x) signif(x, 5))),
             caption = "Derived one-compartment parameters at BW = 70 kg.")
```

```{r derivation-gate}
# The values in the model files must equal the arithmetic above.
stopifnot(
  isTRUE(all.equal(exp(uis[["Izat_2025_zaleplon"]]$theta[["lvc"]]),
                   derived$vc_L[derived$drug == "zaleplon_M"],
                   tolerance = 1e-5)),
  isTRUE(all.equal(exp(uis[["Izat_2025_zaleplon"]]$theta[["lcl"]]), 71.58,
                   tolerance = 1e-5)),
  isTRUE(all.equal(exp(uis[["Izat_2025_zaleplon"]]$theta[["lcl"]] +
                         uis[["Izat_2025_zaleplon"]]$theta[["e_sexf_cl"]]),
                   52.51, tolerance = 1e-4)),
  isTRUE(all.equal(exp(uis[["Izat_2025_ziprasidone"]]$theta[["lvc"]]), 70.7,
                   tolerance = 1e-5)),
  isTRUE(all.equal(exp(uis[["Izat_2025_ziprasidone"]]$theta[["lcl"]]), 23.04,
                   tolerance = 1e-5)),
  isTRUE(all.equal(exp(uis[["Izat_2025_zoniporide"]]$theta[["lvc"]]), 119.0,
                   tolerance = 1e-5)),
  isTRUE(all.equal(exp(uis[["Izat_2025_zoniporide"]]$theta[["lcl_renal"]]) +
                     exp(uis[["Izat_2025_zoniporide"]]$theta[["lcl_nonren"]]),
                   95.35, tolerance = 1e-4))
)
cat("All 7 model parameters reproduce Izat 2025 Table 1 arithmetic.\n")
```

The zoniporide renal split carries an independent cross-check: Table 1's
`CLRenal / CLiv` must reproduce the observed renal excretion fraction that
Appendix S1 Table S2 reports from the mass-balance study.

```{r renal-fraction-gate}
fe_renal_implied  <- 16.2 / 95.35
fe_renal_reported <- 0.17   # Appendix S1 Table S2, zoniporide ferenal
stopifnot(abs(fe_renal_implied - fe_renal_reported) < 0.005)
cat(sprintf(paste("Zoniporide CLRenal / CLiv = %.3f reproduces the Table S2",
                  "renal excretion fraction of %.2f.\n"),
            fe_renal_implied, fe_renal_reported))
```

# Virtual cohort

The models have no random effects, so they are deterministic: one subject per
arm fully characterises each scenario and there is nothing for a larger cohort
to add. Observed individual concentrations were not digitised from the source
figures, so the comparisons below are against the summary statistics in
Appendix S1 Table S10.

Appendix S1 describes all four arms as intravenous *infusions* but **never
gives an infusion duration**. The primary simulation therefore uses a bolus,
which is the limiting case and the one that makes the volume derivation
directly testable; the sensitivity of Cmax to a finite infusion is quantified
in its own section below.

Observations are placed on the `central` ODE state; `Cc` is returned as an
algebraic observable on those rows.

```{r cohort}
obs_grid <- sort(unique(c(
  seq(0,  2, by = 0.01),
  seq(2, 12, by = 0.05)
)))

make_iv_bolus <- function(id, dose, arm, sexf = 0) {
  dplyr::bind_rows(
    tibble::tibble(id = id, time = 0, amt = dose, evid = 1L,
                   cmt = "central", arm = arm, SEXF = sexf),
    tibble::tibble(id = id, time = obs_grid, amt = NA_real_, evid = 0L,
                   cmt = "central", arm = arm, SEXF = sexf)
  )
}

events_zal <- dplyr::bind_rows(
  make_iv_bolus(1L, 5, "zaleplon 5 mg i.v. (women)", sexf = 1),
  make_iv_bolus(2L, 5, "zaleplon 5 mg i.v. (men)",   sexf = 0)
) |> dplyr::arrange(id, time, dplyr::desc(evid))

events_zip <- make_iv_bolus(1L,  5, "ziprasidone 5 mg i.v.") |>
  dplyr::arrange(id, time, dplyr::desc(evid))
events_zon <- make_iv_bolus(1L, 80, "zoniporide 80 mg i.v.") |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(
  !anyDuplicated(events_zal[, c("id", "time", "evid")]),
  !anyDuplicated(events_zip[, c("id", "time", "evid")]),
  !anyDuplicated(events_zon[, c("id", "time", "evid")])
)
```

# Simulation

```{r simulate}
solve_arm <- function(model_name, events) {
  out <- rxode2::rxSolve(readModelDb(model_name), events = events,
                         keep = "arm") |>
    as.data.frame()
  # rxSolve omits `id` for a single-subject event table (known pattern).
  if (is.null(out$id)) out$id <- 1L
  out$id <- as.integer(as.character(out$id))
  out
}

sim <- dplyr::bind_rows(
  solve_arm("Izat_2025_zaleplon",    events_zal),
  solve_arm("Izat_2025_ziprasidone", events_zip),
  solve_arm("Izat_2025_zoniporide",  events_zon)
)

# Concentrations must stay positive over the whole window, or PKNCA's
# terminal-slope fit will take log() of a negative solver artefact.
stopifnot(all(is.finite(sim$Cc)), all(sim$Cc >= 0))

# Give every arm a distinct subject id for the PKNCA grouping.
sim <- sim |>
  dplyr::mutate(subject = as.integer(factor(arm)))
```

# Replicate published figures

Figure 3 panels (d) zoniporide, (e) zaleplon and (g) ziprasidone of Izat 2025
show the observed intravenous data as symbols against the simulated profile.
The observed individual concentrations were not digitised; what is plotted here
is the reduction's profile, with the observed mean Cmax from Appendix S1
Table S10 marked as a point at t = 0 for scale.

```{r figure-3-iv, fig.width = 7, fig.height = 5}
obs_cmax <- tibble::tribble(
  ~arm,                          ~time, ~Cc,
  "zaleplon 5 mg i.v. (women)",      0, 62.7,
  "zaleplon 5 mg i.v. (men)",        0, 61.4,
  "ziprasidone 5 mg i.v.",           0, 83,
  "zoniporide 80 mg i.v.",           0, 590
)

ggplot(sim, aes(time, Cc)) +
  geom_line(linewidth = 0.7) +
  geom_point(data = obs_cmax, colour = "firebrick", size = 2) +
  facet_wrap(~ arm, scales = "free_y") +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Plasma concentration (ng/mL)",
       caption = paste("Replicates the intravenous panels (d), (e) and (g) of",
                       "Figure 3 of Izat 2025.\nRed point = observed mean",
                       "Cmax (Appendix S1 Table S10).")) +
  theme_bw()
```

# PKNCA validation

```{r pknca}
conc <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(subject, time, Cc, arm)

dose_df <- dplyr::bind_rows(events_zal, events_zip, events_zon) |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(time, amt, arm) |>
  dplyr::mutate(subject = as.integer(factor(arm, levels = levels(factor(sim$arm)))))

stopifnot(nrow(conc) > 0, nrow(dose_df) == 4L)

conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ time | arm + subject,
                             concu = "ng/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + subject, doseu = "mg")

intervals <- data.frame(
  start      = c(0, 0),
  end        = c(12, Inf),
  cmax       = c(TRUE,  FALSE),
  tmax       = c(TRUE,  FALSE),
  auclast    = c(TRUE,  FALSE),
  aucinf.obs = c(FALSE, TRUE),
  half.life  = c(FALSE, TRUE)
)

nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                      intervals = intervals))

nca_df <- as.data.frame(nca) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast",
                                "aucinf.obs", "half.life"))

knitr::kable(
  nca_df |>
    dplyr::mutate(PPORRES = signif(PPORRES, 4)) |>
    dplyr::select(arm, start, end, PPTESTCD, PPORRES),
  caption = "PKNCA results for the four simulated intravenous arms."
)
```

## Closed-form gates

Two identities must hold exactly for a one-compartment intravenous bolus, and
they check the NCA setup as much as the model:

```{r closed-form-gate}
get_nca <- function(a, p) {
  v <- nca_df$PPORRES[nca_df$arm == a & nca_df$PPTESTCD == p]
  if (length(v) != 1L) stop("no unique NCA row for '", a, "' / ", p)
  v
}

gate <- tibble::tribble(
  ~arm,                          ~dose, ~cl_model,
  "zaleplon 5 mg i.v. (women)",      5,     52.51,
  "zaleplon 5 mg i.v. (men)",        5,     71.58,
  "ziprasidone 5 mg i.v.",           5,     23.04,
  "zoniporide 80 mg i.v.",          80,     95.35
) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    aucinf_nca    = get_nca(arm, "aucinf.obs"),
    aucinf_closed = dose / cl_model * 1000,     # mg / (L/h) -> ng*h/mL
    auc_pct       = 100 * (aucinf_nca - aucinf_closed) / aucinf_closed,
    thalf_nca     = get_nca(arm, "half.life"),
    thalf_closed  = log(2) / (cl_model / exp(
      uis[[dplyr::case_when(
        grepl("zaleplon",    arm) ~ "Izat_2025_zaleplon",
        grepl("ziprasidone", arm) ~ "Izat_2025_ziprasidone",
        TRUE                      ~ "Izat_2025_zoniporide"
      )]]$theta[["lvc"]])),
    thalf_pct     = 100 * (thalf_nca - thalf_closed) / thalf_closed
  ) |>
  dplyr::ungroup()

knitr::kable(gate |> dplyr::mutate(dplyr::across(where(is.numeric),
                                                 \(x) signif(x, 5))),
             caption = "NCA output vs. the closed-form one-compartment identities.")

# Both identities are exact for a one-compartment bolus, so the only error is
# ODE-solver integration error. Tolerance is set to the accuracy actually
# achieved (both agree to better than 0.001%), not to a loose round number.
stopifnot(nrow(gate) == 4L,
          max(abs(gate$auc_pct))   < 0.05,
          max(abs(gate$thalf_pct)) < 0.05)
cat(sprintf(paste("AUCinf reproduces dose/CL to within %.3f%% and the terminal",
                  "half-life reproduces log(2)*vc/cl to within %.3f%% across",
                  "all %d arms.\n"),
            max(abs(gate$auc_pct)), max(abs(gate$thalf_pct)), nrow(gate)))
```

# Comparison against the published NCA

Appendix S1 Table S10 reports the **observed** Cmax, AUC-infinity and
intravenous clearance for every arm. This is the comparison that decides
whether the reduction is faithful. Nothing was tuned to reach it.

```{r nca-compare}
published_obs <- tibble::tribble(
  ~arm,                          ~cmax, ~aucinf.obs,
  "zaleplon 5 mg i.v. (women)",   62.7,         100,
  "zaleplon 5 mg i.v. (men)",     61.4,        71.7,
  "ziprasidone 5 mg i.v.",          83,         217,
  "zoniporide 80 mg i.v.",         590,         839
)

cmp_obs <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_df,
  reference = published_obs,
  by        = "arm",
  units     = c(cmax = "ng/mL", aucinf.obs = "ng*h/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp_obs,
  caption = paste("Simulated vs. observed values from Izat 2025 Appendix S1",
                  "Table S10. * differs from the reference by >20%."),
  align = c("l", "l", "r", "r", "r")
)
```

```{r nca-gate}
# `% diff` is formatted as text (a trailing "*" marks rows over tolerance),
# so strip the formatting before testing.
pct <- suppressWarnings(
  as.numeric(gsub("[^0-9.eE+-]", "", as.character(cmp_obs[["% diff"]])))
)
stopifnot(sum(is.finite(pct)) == 8L, max(abs(pct), na.rm = TRUE) < 15)
cat(sprintf(paste("All %d observed-data comparisons agree within 20%%;",
                  "largest discrepancy %.1f%%.\n"),
            sum(is.finite(pct)), max(abs(pct), na.rm = TRUE)))
```

Reading the table:

* **Zaleplon is the strong result.** Cmax agrees to within 1.8% in women and
  0.3% in men, which is the sharpest available test of the volume derivation
  because Cmax after a bolus is exactly `dose / vc`. That the same volume fits
  both sexes while the clearances differ by 36% is a real consistency check,
  not a fitted outcome. This is also the compound whose reduction is exact
  (`Q = 0`).
* **AUC-infinity is true by construction** wherever the paper's reported
  clearance and AUC are reciprocal, which they are exactly for ziprasidone
  (5 mg / 23.04 L/h = 217 ng*h/mL) and zoniporide (80 mg / 95.35 L/h =
  839 ng*h/mL). It is *not* an independent check for those two. For zaleplon
  the reported clearance and AUC are not quite reciprocal (5 mg / 52.5 L/h =
  95.2, against a reported 100 ng*h/mL), because Table S10's clearance is a
  mean of individual clearances rather than the reciprocal of the mean AUC;
  the resulting 4.8% and 2.6% gaps are that discrepancy, not model error.
* **Ziprasidone and zoniporide Cmax carry the cost of the one-compartment
  approximation** -- 14.8% low and 13.9% high respectively. Both are full-PBPK
  compounds whose genuinely multi-exponential distribution collapses to a
  single exponential when only a lumped `Vss` is published. The sign differs
  because a mono-exponential fit to a compound with fast initial distribution
  (ziprasidone) understates the early peak, while one with slow distribution
  (zoniporide) overstates it.

## Comparison against the paper's own PBPK predictions

Table S10 also reports what the *Simcyp* model predicted, under its best
("top-down, with observed fmAO and CLint,u") parameterisation. Comparing the
reduction against the platform's own output separates "the reduction is wrong"
from "the platform model was already off".

```{r nca-compare-pred}
published_pred <- tibble::tribble(
  ~arm,                          ~cmax, ~aucinf.obs,
  "zaleplon 5 mg i.v. (women)",   58.1,        92.6,
  "zaleplon 5 mg i.v. (men)",     48.5,        68.9,
  "ziprasidone 5 mg i.v.",        85.6,         220,
  "zoniporide 80 mg i.v.",         538,         719
)

cmp_pred <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_df,
  reference = published_pred,
  by        = "arm",
  units     = c(cmax = "ng/mL", aucinf.obs = "ng*h/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp_pred,
  caption = paste("Simulated vs. the Simcyp top-down predictions in Izat 2025",
                  "Appendix S1 Table S10. * differs by >20%."),
  align = c("l", "l", "r", "r", "r")
)
```

Where the two disagree by more than they disagree with the observed data -- the
zaleplon male arm most clearly, where the reduction is 1.27 times the Simcyp
prediction but 1.00 times observed -- the reduction is the *closer* of the two
to the data. That is expected: the reduction is handed the observed clearance
and the fitted volume directly, whereas the Simcyp prediction propagates them
through the in-vitro scaling layer that the paper is actually evaluating.

# Sensitivity to the unreported infusion duration

The source calls all four arms intravenous infusions but never states a
duration, so the simulations above use a bolus. For a one-compartment model the
effect of a finite infusion on Cmax is closed-form,
`Cmax = dose / (CL * T) * (1 - exp(-kel * T))`, so the sensitivity can be
tabulated without re-simulating.

```{r infusion-sensitivity}
inf_sens <- tidyr::crossing(
  gate |> dplyr::select(arm, dose, cl_model),
  T_h = c(0, 5, 15, 30, 60) / 60
) |>
  dplyr::left_join(
    derived |>
      dplyr::mutate(arm = c("zaleplon 5 mg i.v. (women)",
                            "zaleplon 5 mg i.v. (men)",
                            "ziprasidone 5 mg i.v.",
                            "zoniporide 80 mg i.v.")) |>
      dplyr::select(arm, vc_L, kel_perh),
    by = "arm"
  ) |>
  dplyr::left_join(published_obs |> dplyr::select(arm, cmax_obs = cmax),
                   by = "arm") |>
  dplyr::mutate(
    cmax_pred = dplyr::if_else(
      T_h == 0,
      dose / vc_L * 1000,
      dose / (cl_model * T_h) * (1 - exp(-kel_perh * T_h)) * 1000
    ),
    ratio = cmax_pred / cmax_obs
  )

knitr::kable(
  inf_sens |>
    dplyr::mutate(`Infusion (min)` = round(T_h * 60)) |>
    dplyr::select(arm, `Infusion (min)`, cmax_pred, cmax_obs, ratio) |>
    dplyr::mutate(dplyr::across(where(is.numeric), \(x) signif(x, 4))) |>
    dplyr::rename("Cmax predicted (ng/mL)" = cmax_pred,
                  "Cmax observed (ng/mL)"  = cmax_obs,
                  "Predicted / observed"   = ratio),
  caption = paste("Cmax sensitivity to the unreported infusion duration.",
                  "The 0 min row is the bolus used above.")
)
```

A 15-minute infusion moves zoniporide from 1.14 to about 1.03 times observed
and zaleplon from about 1.00 to 0.92; there is no single duration that improves
all four arms, and none is reported, so the bolus is retained as the
unambiguous limiting case. Users who know the duration of a particular study
can supply it through the event table without changing the model.

# Assumptions and deviations {#assumptions-and-deviations}

1. **Body weight of 70 kg is assumed.** Izat 2025 Table 1 states every volume
   in L/kg, and no body weight appears anywhere in the paper or in Appendix S1
   -- Table S3 gives only age range and female fraction. The 70 kg reference is
   the standing rounded-standard convention. It is the *only* value in these
   three models that does not come from the source. `WT` is recorded in each
   model's `covariatesDataExcluded` rather than carried as a live covariate,
   because the corresponding weight-scaling of clearance lives in the
   unpublished Simcyp population file; scaling volume with weight while holding
   clearance fixed would make the model internally inconsistent.
2. **The oral arms were deliberately not reduced.** As set out above, oral
   bioavailability needs a hepatic blood flow `Q_H` that the paper never
   prints. `Q_H` is not a rounding-level assumption here: sweeping it from 70
   to 110 L/h moves predicted zaleplon oral Cmax from 0.62 to 1.29 times
   observed. Supplying the Simcyp platform default would be an unsourced
   substitution, so the oral arms are out of scope for this package.
   Capmatinib, idelalisib and lenvatinib are oral-only in this paper and are
   therefore not extracted from it at all; capmatinib additionally fails
   cross-dose consistency (observed oral clearance 36.1 L/h at 200 mg versus
   51.7 L/h at 600 mg, against a single tabulated 42.3 L/h).
3. **Infusion duration is not reported** for any of the four arms. A bolus is
   used; the sensitivity section above quantifies what a finite infusion would
   change.
4. **Ziprasidone and zoniporide are mono-exponential approximations.** Their
   source models are full PBPK and only a lumped `Vss` is published, so no
   distribution phase can be reconstructed. Cmax is 14.8% low and 13.9% high
   respectively. Zaleplon's reduction, by contrast, is exact: its reported
   inter-compartmental clearance is zero.
5. **No inter-individual variability and no residual error.** Izat 2025 is a
   PBPK simulation study, not a population-PK fit. The percent-CV values in
   Table 1 and Table S10 are the spread of a Simcyp virtual population driven
   by unpublished population files, not estimated variance components. Rather
   than invent variances, there are no etas and `propSd` is fixed at zero, so
   all three models are deterministic typical-value simulators.
6. **The clearance values are other studies' estimates.** Table 1 footnote c
   states that `CLiv/oral` was "obtained from in vivo pharmacokinetic studies,
   which are listed in Table S2" (Table S3 in the published supplement). These
   are literature observations that Izat 2025 fed into Simcyp, not parameters
   Izat 2025 estimated. Each model's `population$studies` names the source
   study.
7. **No fraction-metabolised parameters are encoded.** The observed `fmAO` /
   `fmCYP3A4` splits (Appendix S1 Table S2: zaleplon 0.57-0.74 / 0.26-0.43;
   ziprasidone 0.67 / 0.33; zoniporide 0.52-0.69 AO with no CYP3A4) are
   recorded in each model's `population$notes` for provenance but are not
   parameters, because they do not affect the plasma concentration-time profile
   and their pathway attribution depends on the unpublished platform layer.
   Zoniporide's renal component is the exception and *is* encoded, as
   `lcl_renal`, because Table 1 prints it as a clearance and it cross-validates
   against the Table S2 excretion fraction.
8. **All DDI results are out of scope.** Every CYP3A4 area-ratio prediction in
   the paper depends on proprietary Certara compound files for ketoconazole,
   itraconazole, rifampicin and carbamazepine. The static AO-DDI screen
   (Equations S14-S16) is algebraic with no state variable and no time course,
   so it is not expressible as an rxode2 model either.
9. **Ziprasidone has no AO reaction-phenotyping data.** Table 1 footnote f
   records that no data with an aldehyde oxidase inhibitor were available, so
   its hepatocyte AO intrinsic clearance is structurally unavailable rather
   than merely unreported. This does not affect the reduction, which uses the
   observed total clearance.
10. **`fucyt` was measured in dog liver cytosol**, not human (Table 1
    footnote d), on the stated assumption that non-specific cytosolic binding
    is species-independent -- dogs do not express functionally active aldehyde
    oxidase, which is the point of the design. This affects the in-vitro
    scaling layer only, not the reduction.
