---
title: "Jager_2011_gemtuzumab"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Jager_2011_gemtuzumab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Jager E, van der Velden VHJ, te Marvelde JG, Walter RB, Agur Z,
  Vainstein V (2011). *Targeted drug delivery by gemtuzumab ozogamicin:
  mechanism-based mathematical model for treatment strategy improvement and
  therapy individualization.* PLoS One 6(9):e24265.
  [doi:10.1371/journal.pone.0024265](https://doi.org/10.1371/journal.pone.0024265).
* DDMORE Foundation Model Repository entry:
  [DDMODEL00000229](https://repository.ddmore.eu/model/DDMODEL00000229)
  -- Monolix re-fit with an added peripheral drug compartment.
* Description: Mechanism-based PKPD model coupling drug pharmacokinetics with
  explicit binding to the cell-surface antigen CD33. Free drug in a central
  compartment binds free CD33 receptor to form a drug-receptor complex that is
  internalized; the toxic ozogamicin component then drives linear depletion of
  leukemic blast cells.

The model packaged here translates the DDMORE bundle's `GO_PK_model.mdl`
(MDL DSL source) to the nlmixr2lib `inst/modeldb/ddmore/` convention.
Parameter values come from the bundle's MDL `parObj$STRUCTURAL` and
`parObj$VARIABILITY` blocks, which `Model_Accommodations.txt` documents as
the Monolix-fitted final estimates obtained by re-fitting the Jager 2011
mechanism on the original publication data with an additional peripheral
PK compartment and simultaneous (rather than step-wise) estimation of all
parameters. The DDMORE bundle does **not** ship a `.lst` listing or a
simulated dataset, and the original Jager 2011 PLoS One paper is not on
disk under this worktree's literature tree, so the implementation is
faithful to the MDL but has no external numeric cross-check; see
**Assumptions and deviations** below.

# Population

The Jager 2011 publication describes a single-cohort mechanism-based PKPD
analysis in patients with acute myeloid leukemia (AML) receiving the
CD33-directed antibody-drug conjugate gemtuzumab ozogamicin (Mylotarg).
The DDMORE bundle does not reproduce demographic detail (age, weight,
sex, race distribution, regional balance, or trial design) in either the
`Model_Accommodations.txt` accompaniment or the MDL `dataObj` block.
Programmatic access:

```{r population-meta}
mod_fun <- readModelDb("Jager_2011_gemtuzumab")
str(environment(mod_fun)$population)
```

Consult Jager et al. (2011) directly for the full population description.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in
`inst/modeldb/ddmore/Jager_2011_gemtuzumab.R`. The table below collects
those entries in one place. All values come from the MDL bundle's
`parObj` block in `GO_PK_model.mdl`; the bundle ships no `Output_real_*.lst`
listing.

| nlmixr2 parameter | Value      | DDMORE source location                                              |
|-------------------|------------|---------------------------------------------------------------------|
| `lvc`             | log(5.42)  | `parObj$STRUCTURAL$POP_v1`                                          |
| `lk`              | log(0.0135)| `parObj$STRUCTURAL$POP_k`                                           |
| `lkm`             | log(0.00757)| `parObj$STRUCTURAL$POP_km`                                         |
| `lkn`             | log(0.0185)| `parObj$STRUCTURAL$POP_kn`                                          |
| `lkb`             | log(9.24e5)| `parObj$STRUCTURAL$POP_kb`                                          |
| `lku`             | log(310)   | `parObj$STRUCTURAL$POP_ku`                                          |
| `lki`             | log(0.624) | `parObj$STRUCTURAL$POP_ki`                                          |
| `lke`             | log(0.199) | `parObj$STRUCTURAL$POP_ke`                                          |
| `lrp`             | log(823)   | `parObj$STRUCTURAL$POP_rp`                                          |
| `lalph`           | log(0.0755)| `parObj$STRUCTURAL$POP_alph`                                        |
| `ln0`             | log(1.32e-5)| `parObj$STRUCTURAL$POP_n0`                                         |
| `etalki`          | 0.258^2     | `parObj$VARIABILITY$omega_ki` (sd, type is sd)                      |
| `etalke`          | 0.281^2     | `parObj$VARIABILITY$omega_ke` (sd, type is sd)                      |
| `etalrp`          | 0.611^2     | `parObj$VARIABILITY$omega_rp` (sd, type is sd)                      |
| `addSd`           | 0.000583   | `parObj$STRUCTURAL$a` (additive residual error)                     |
| Constant `nav`    | 6.02214    | `mdlObj$MODEL_PREDICTION` inline (Avogadro number / 1e23)           |
| Constant `vblood` | 4.8        | `mdlObj$MODEL_PREDICTION` inline (assumed blood volume, L)          |
| ODE `central`     | n/a        | `mdlObj$MODEL_PREDICTION$DEQ` `A1`                                  |
| ODE `target`      | n/a        | `mdlObj$MODEL_PREDICTION$DEQ` `A2` (init = rp / ke)                 |
| ODE `complex`     | n/a        | `mdlObj$MODEL_PREDICTION$DEQ` `A3` (init = 0)                       |
| ODE `cells`       | n/a        | `mdlObj$MODEL_PREDICTION$DEQ` `A4` (init = n0)                      |
| ODE `peripheral1` | n/a        | `mdlObj$MODEL_PREDICTION$DEQ` `A5` (init = 0)                       |
| Output `Cc`       | n/a        | `mdlObj$MODEL_PREDICTION$output1 = A1 / v1`                         |

# Virtual cohort

The DDMORE bundle for DDMODEL00000229 ships no event dataset and the
Jager 2011 paper is not on disk, so no published clinical regimen is
available to reproduce. The vignette exercises the packaged ODEs on a
single typical-individual scenario (one 50 mg IV bolus into the central
compartment) so that the simulated trajectories can be inspected for
mechanistic plausibility -- *not* as a clinically-grounded dosing
regimen.

```{r cohort}
events <- data.frame(
  id   = 1L,
  time = c(0, seq(0.25, 336, by = 0.25)),
  amt  = c(50, rep(NA_real_, length(seq(0.25, 336, by = 0.25)))),
  evid = c(1L, rep(0L,  length(seq(0.25, 336, by = 0.25)))),
  cmt  = c("central", rep("central", length(seq(0.25, 336, by = 0.25))))
)
```

# Simulation

```{r simulate}
mod  <- rxode2::rxode2(readModelDb("Jager_2011_gemtuzumab"))
mod_typ <- rxode2::zeroRe(mod)
sim  <- rxode2::rxSolve(mod_typ, events = events)
```

# Mechanistic sanity checks

The DDMORE bundle does not link to a `.lst` listing or a published NCA /
VPC figure that this packaged model can be compared against numerically,
and the structural model has no NCA-amenable PK arm in the ordinary sense
(drug elimination is split across linear elimination, target-mediated
internalization, and a peripheral PK compartment with two rate constants
that are not in CL/Q form). Validation therefore consists of mechanistic
sanity checks: each ODE has the qualitative behavior its mechanism implies.

## Drug central concentration declines monotonically after the bolus

```{r mech-cc}
ggplot(sim, aes(time, Cc)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Cc (mg/L; assumed unit, see Errata)",
       title = "Central drug concentration")
```

```{r mech-cc-check}
stopifnot(sim$Cc[1] > 8)               # initial ~= amt / vc = 50 / 5.42 ~= 9.2
stopifnot(all(diff(sim$Cc) < 0))       # monotonic decline (no rebound)
stopifnot(tail(sim$Cc, 1) > 0)
```

## Free CD33 receptor depletes upon drug arrival, then recovers as drug clears

The receptor is at steady state `rp / ke ~= 4135` at `t = 0`, but the
incoming dose drives a fast binding pulse that depletes free target by
several orders of magnitude before `target` slowly recovers as drug --
and complex -- fall.

```{r mech-target}
ggplot(sim, aes(time, target)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Free CD33 target (relative units)",
       title = "Free target dynamics")
```

```{r mech-target-check}
target_steady <- with(environment(mod_fun), exp(log(823) - log(0.199)))  # rp / ke
stopifnot(min(sim$target) < 1)                       # transient near-zero depletion
stopifnot(tail(sim$target, 1) > min(sim$target) * 5) # partial recovery
stopifnot(target_steady > 4000 && target_steady < 4200)
```

## Drug-receptor complex peaks early and decays as drug clears

```{r mech-complex}
ggplot(sim, aes(time, complex)) +
  geom_line() +
  labs(x = "Time (h)", y = "Drug-receptor complex (relative units)",
       title = "Complex dynamics")
```

```{r mech-complex-check}
stopifnot(sim$complex[1] > 0)
stopifnot(max(sim$complex) > 1000)
peak_idx  <- which.max(sim$complex)
final_idx <- nrow(sim)
stopifnot(sim$complex[final_idx] < sim$complex[peak_idx])
```

## Leukemic blast cells decline at the typical-population rate `alph`

The cell ODE is `d/dt(cells) = -alph * cells`, so cells decay
exponentially at typical-value rate `alph = 0.0755 1/h` (half-life
`log(2) / alph ~= 9.18 h`). The simulated trajectory should match this
analytically.

```{r mech-cells}
ggplot(sim, aes(time, cells)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Leukemic blasts (relative units)",
       title = "Blast cell depletion")
```

```{r mech-cells-check}
alph_typ <- 0.0755
expected <- 1.32e-5 * exp(-alph_typ * sim$time)
rel_err  <- abs(sim$cells - expected) / expected
stopifnot(max(rel_err) < 0.001)  # within 0.1% of the analytical solution
```

## Drug peripheral compartment fills then re-distributes

```{r mech-peri}
ggplot(sim, aes(time, peripheral1)) +
  geom_line() +
  labs(x = "Time (h)", y = "Drug in peripheral1 (mg)",
       title = "Peripheral drug compartment")
```

```{r mech-peri-check}
stopifnot(sim$peripheral1[1] > 0)
stopifnot(any(diff(sim$peripheral1) > 0))      # at least transient accumulation
stopifnot(any(diff(sim$peripheral1) < 0))      # then re-distribution / decline
```

# Assumptions and deviations

The DDMORE Foundation Model Repository entry DDMODEL00000229 ships only
the MDL DSL source (`GO_PK_model.mdl`), its PharmML XML rendering
(`GO_PK_model.xml`), the `Model_Accommodations.txt` reference note, and
RDF / scraper metadata -- there is **no** `Output_real_*.lst`,
`Output_simulated_*.lst`, or simulated event dataset. The Jager 2011
publication itself is not on disk in this worktree's literature tree.
The following assumptions and deviations apply.

* **Parameter values come from the MDL `parObj` block, not a `.lst`
  listing.** The bundle has no NONMEM listing file with a
  `MINIMIZATION SUCCESSFUL` block whose final estimates can be read off,
  and the DDMORE-source skill's default rule (read final estimates from
  `Output_real_*.lst`) cannot apply. The MDL `parObj$STRUCTURAL` and
  `parObj$VARIABILITY` values are treated as the bundle's authoritative
  final estimates per `Model_Accommodations.txt`, which describes the
  parObj as the Monolix re-fit of the Jager 2011 model on the
  publication's data.
* **Bundle is a Monolix re-fit, not a literal reproduction of Jager 2011.**
  `Model_Accommodations.txt` documents two material differences from
  the publication:
  1. The PK structure was extended from the published one-compartment
     drug PK (linear elimination only) to a two-compartment drug PK
     with linear elimination plus a peripheral compartment (rate
     constants `km`, `kn`).
  2. All parameters were estimated simultaneously in Monolix, in
     contrast with Jager 2011's step-wise estimation. As a result, the
     parObj values for parameters that exist in both formulations
     (e.g., `kb`, `ku`, `ki`, `ke`, `rp`, `k`, `alph`, `n0`, `v1`)
     differ from the values reported in the original publication.

  The packaged model reproduces the DDMORE bundle's Monolix-re-fit
  formulation, **not** the published Jager 2011 formulation.
* **No external numeric cross-check.** Without the publication PDF on
  disk and without a `.lst` listing in the bundle, neither published
  parameter tables nor the bundle's own re-fit objective function /
  per-subject predictions are available for verification. Source-trace
  comments point at the MDL `parObj` lines; reviewers wishing to verify
  against the original Monolix run should consult the live DDMORE
  repository entry at
  `https://repository.ddmore.eu/model/DDMODEL00000229`.
* **Units of dose, central concentration, additive error, receptor, and
  cell counts are MDL-native.** The MDL `dataObj` block declares
  `AMT : { use is amt, variable = A1 }` and `DV : { use is dv, variable = Y }`
  with **no** explicit unit annotations. The packaged `units` field
  declares `dosing = "mg"` and `concentration = "mg/L"` based on the
  observation that a 50-unit dose into `central` produces an initial
  `Cc = amt / vc ~= 9.22`, which is consistent with a clinical mg-scale
  gemtuzumab ozogamicin bolus (typical adult IV ~6-18 mg) and a
  central volume `v1 = 5.42 L`. Receptor units (`target`, `complex`,
  `rp`) and cell-count units (`cells`, `n0`) are MDL-native and are
  mechanism-specific scalings rather than physical receptor or cell
  numbers; the inline constants `nav = 6.02214` (Avogadro / 1e23) and
  `vblood = 4.8 L` in the `central` ODE convert the per-cell binding
  flux into a plasma-equivalent rate. Reviewers concerned about the
  exact unit interpretation should consult Jager 2011 directly.
* **OMEGA correlation block is dropped as vestigial.** The MDL parObj
  declares an `OMEGA` block with `type is corr` over six etas
  (`ETA_k`, `ETA_alph`, `ETA_n0`, `ETA_v1`, `ETA_kn`, `ETA_km`), but
  every individual `omega_X` for those six parameters is `0.0` (`type is sd`).
  Mathematically, a correlation among zero-variance random effects has
  no effect on simulation; the packaged model omits the correlation
  block and applies IIV only to `ki`, `ke`, and `rp` (the three etas
  whose `omega_X` parObj sd is non-zero).
* **Compartment name `cells` is non-canonical.**
  `nlmixr2lib::checkModelConventions()` flags `cells` as not in the
  canonical compartment list. The model's biology is leukemic blast
  cell count, which has no canonical short name in the register;
  `cells` was chosen by analogy with the `circ` / `precursor1..4`
  naming used in the DDMORE bundle for `Friberg_2002_paclitaxel`. A
  GitHub issue should ratify a canonical name for cell-count
  compartments before this convention is reused for additional models.
* **Population demographics absent.** The `population` metadata for
  this model uses `NA_*` placeholders for age range, weight range, sex
  ratio, region, and study count because the DDMORE bundle does not
  ship that detail and the publication itself is not on disk. Consult
  Jager 2011 for the population description.
* **No PKNCA validation.** The structural model is not a standard PK
  model, lacks NCA-amenable single-arm PK output for which Cmax / AUC
  / half-life are clinically interpretable in isolation, and has no
  published reference NCA values or VPC figure available locally to
  compare against. Validation in this vignette is reduced to
  mechanistic sanity (each ODE behaves as its mechanism implies) and
  an analytical match for the leukemic-blast depletion ODE.
