---
title: "Dordaviprone (Jaiswal 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dordaviprone (Jaiswal 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Jaiswal_2025_dordaviprone"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Article: <https://doi.org/10.1002/psp4.70093>\n")
cat("* Supplement (Data S1, Tables S1-S8): ",
    "<https://www.ebi.ac.uk/europepmc/webservices/rest/PMC12521050/supplementaryFiles>\n",
    sep = "")
```

Dordaviprone (ONC201) is a brain-penetrant imipridone developed for H3 K27M-mutant
glioma. Jaiswal 2025 built a **minimal distribution PBPK model with a single
adjusting compartment (SAC) and mechanistic ADAM absorption**, in the Simcyp
Population-based Simulator (V21), to assess CYP3A4 drug-interaction risk.

## What this model file is, and what it is not

This is a **reduction** of the published Simcyp model, not a port of it. That
distinction matters for how the numbers below should be read.

What is reproducible, and is encoded here:

* the minimal-PBPK disposition (systemic compartment exchanging with the SAC via
  the paper's `k_in` / `k_out`);
* concentration-dependent clearance driven by saturable alpha-1-acid glycoprotein
  binding, from the five-point `fu` table in Table 1;
* first-pass extraction split into gut-wall and hepatic components;
* the six CYP3A4-modulator arms, each entering through a single relative-CYP3A4
  activity term.

What is **not** reproducible from published information, and is therefore absent:

* the Simcyp whole-body mass-balance equations and the ADAM gut model (neither is
  published), so the food effect and the proton-pump-inhibitor arm -- both of which
  act inside ADAM -- are not encoded. Both published effects on AUC are essentially
  null, so the no-modulator model reproduces those arms' exposures anyway;
* the perpetrator predictions (dordaviprone inhibiting CYP3A4 / CYP2C8 / CYP2D6),
  which need proprietary Simcyp *substrate* compound files. The paper's result there
  is a null one (all ratios 1.00);
* the Simcyp virtual-population variability (about 62% CV on AUC, 45% on Cmax),
  which is driven by unpublished demographic and enzyme-abundance distributions.
  This model is therefore a **typical-value** model: no etas, and `propSd` fixed at 0.

Two parameters could not be transcribed and were **calibrated**; they are called out
explicitly in "Assumptions and deviations" at the end.

# Population

```{r population, echo = FALSE, results = "asis"}
pop <- ui$population
cat("* Species: ", pop$species, "\n", sep = "")
cat("* Subjects: ", pop$n_subjects, " (across ", pop$n_studies, " clinical studies)\n", sep = "")
cat("* Doses: ", pop$dose_range, "\n", sep = "")
cat("* Region / virtual population: ", pop$regions, "\n", sep = "")
```

PK data came from four phase I studies in healthy participants (Section 2.2):
ONC201-101 Part A single ascending dose (n = 15; the 625 mg arm was used for model
*development*, the 125 and 375 mg arms for *verification*); Part B2, the high-fat-meal
arm (n = 30); the `[14C]`-dordaviprone mass-balance study (n = 6), which fixed the
fraction absorbed and the zero renal clearance; clinical study 103, the itraconazole
DDI study (n = 18), whose observed CL/F of 29 L/h anchors the retrograde clearance
calculation and whose AUC and Cmax ratios calibrated fm CYP3A4; and clinical study
107 (n = 16), the rabeprazole arm.

No patient data were used. The paper argues the same magnitudes of change would apply
in the target glioma population (Discussion), but that is an extrapolation.

The same information is available programmatically via
`readModelDb("Jaiswal_2025_dordaviprone")()$population`.

# Source trace

Every `ini()` entry carries an in-file comment naming its origin. Collected here for
review. "Table 1" and "Table 2" are the main paper; "Table S*n*" is Data S1.

| Parameter | Value | Source |
|---|---|---|
| `lka` | 1.580 1/h | **Calibrated** to the simulated median Tmax of 1.20 h at 625 mg (Table S1). The source used ADAM and reports no first-order ka. Cross-check: Table 1 `P_eff,man` 3.74e-04 cm/s implies ka = 2*Peff/r = 1.795 1/h at r = 1.5 cm, within 12%. |
| `lvc` | 143.20 L | Derived: Table 1 Vss 1.11 L/kg and V_SAC 6.60e-05 L/kg give Vsys = (1.11 - 0.000066)*70 - 1.65 = 76.045 L; divided by the baseline bioavailability 0.531058 to give an apparent volume. |
| `lk12` | 0.11 1/h | Table 1 `k_in`, fitted by the authors to the 625 mg profile. |
| `lk21` | 0.15 1/h | Table 1 `k_out`. |
| `lclint` | 968.18 L/h | Derived so that CL/F returns the Table 1 observed value of 29 L/h at 125 mg. |
| `fm_cyp3a4` | 0.8 | Table 1 `fm CYP3A4` (remainder 0.2 to CYP2D6). |
| `fumin`, `fumax`, `cup50` | 0.0283, 0.0628, 5.337 mg/L | Fitted to the five printed `fu`-vs-concentration points in Table 1. |
| `eh`, `egut` | 0.2199, 0.1490 | Jointly pinned by the printed CL/F = 29 L/h (with f_a 0.8, B:P 0.778, Q_H 90 L/h) and the calibrated baseline bioavailability. |
| `e_conmed_itraconazole_cyp3a4` | log(0.1378) | Back-solved from the Table S6 AUC ratio 4.62 (125 mg). |
| `e_conmed_erythromycin_cyp3a4` | log(0.3722) | Back-solved from the AUC ratio 2.68 (Abstract; Figure 3). |
| `e_conmed_fluconazole_cyp3a4` | log(0.4102) | Back-solved from the AUC ratio 2.48 (Abstract; Figure 3). |
| `e_conmed_cimetidine_cyp3a4` | log(0.7386) | Back-solved from the AUC ratio 1.42 (Abstract; Figure 3). |
| `e_conmed_efv_cyp3a4` | log(2.0791) | Back-solved from the AUC ratio 0.349 (Figure 3; Abstract and Discussion round to 0.35). |
| `e_conmed_rifampicin_cyp3a4` | log(3.1347) | Back-solved from the AUC ratio 0.167 (Figure 3; Abstract and Discussion round to 0.17). |
| `propSd` | fixed(0) | No residual-error model is reported; this is a simulation model. |
| ODE structure | n/a | Figure 1 (modeling-strategy schematic) and Section 2.3. |

Un-printed standard physiological constants used in the derivations: body weight 70 kg,
liver volume 1.65 L, hepatic blood flow 90 L/h, small-intestinal radius 1.5 cm (the last
only for the ka cross-check). Varying hepatic blood flow by +/-20% moves the maximum
attainable bioavailability increase by only 3%, so nothing DDI-related is sensitive to it.

# Saturable protein binding

The load-bearing piece of this reduction. Table 1 tabulates `fu` at five total plasma
concentrations; the model carries a saturable form fitted to those five points.

```{r fu-check}
ini_vals <- ui$theta
fu_of <- function(cconc) {
  ini_vals[["fumin"]] +
    (ini_vals[["fumax"]] - ini_vals[["fumin"]]) * cconc / (ini_vals[["cup50"]] + cconc)
}

fu_tab <- tibble::tribble(
  ~conc_mg_L, ~fu_published,
  0.193,      0.0294,
  0.386,      0.0303,
  1.932,      0.0389,
  3.865,      0.0412,
  7.730,      0.0492
) |>
  dplyr::mutate(
    fu_model = fu_of(conc_mg_L),
    pct_diff = 100 * (fu_model / fu_published - 1)
  )

# The fit must reproduce every printed point closely; 4% is the accuracy achieved.
stopifnot(nrow(fu_tab) == 5L, max(abs(fu_tab$pct_diff)) < 4)

fu_tab |>
  dplyr::rename(
    "Total plasma conc (mg/L)" = conc_mg_L,
    "fu (Table 1)"             = fu_published,
    "fu (model)"               = fu_model,
    "% difference"             = pct_diff
  ) |>
  knitr::kable(digits = c(3, 4, 4, 2),
               caption = "Saturable AAG binding: model vs the five printed points of Jaiswal 2025 Table 1.")
```

Because the hepatic availability cancels between numerator and denominator,
`CL/F = (CLint/F) * fu` exactly. So this one relationship is what makes apparent
clearance rise with dose, which is checked numerically further below.

# Virtual cohort and simulation

The model carries no random effects, so a "cohort" is one representative subject per
arm; simulating more would return identical profiles. Ten arms are simulated: three
dose levels alone, and the six CYP3A4-modulator arms against their matched controls.

```{r cohort}
set.seed(20250820)

conmed_cols <- c("CONMED_ITRACONAZOLE", "CONMED_ERYTHROMYCIN", "CONMED_FLUCONAZOLE",
                 "CONMED_CIMETIDINE", "CONMED_EFV", "CONMED_RIFAMPICIN")

arms <- tibble::tribble(
  ~arm,                        ~dose, ~modulator,             ~tend,
  "125 mg alone",                125, NA_character_,            168,
  "375 mg alone",                375, NA_character_,            168,
  "625 mg alone",                625, NA_character_,            168,
  "125 mg alone (DDI control)",  125, NA_character_,            336,
  "125 mg + itraconazole",       125, "CONMED_ITRACONAZOLE",    336,
  "625 mg alone (DDI control)",  625, NA_character_,            336,
  "625 mg + erythromycin",       625, "CONMED_ERYTHROMYCIN",    336,
  "625 mg + fluconazole",        625, "CONMED_FLUCONAZOLE",     336,
  "625 mg + cimetidine",         625, "CONMED_CIMETIDINE",      336,
  "625 mg + efavirenz",          625, "CONMED_EFV",             336,
  "625 mg + rifampicin",         625, "CONMED_RIFAMPICIN",      336
)

# One event table per arm. Observations go on the `central` ODE state; rxode2
# returns the algebraic observable Cc as a column at those rows.
make_arm <- function(i) {
  a <- arms[i, ]
  tgrid <- sort(unique(c(seq(0, 12, by = 0.05), seq(12, a$tend, by = 0.25))))
  ev <- rxode2::et(amt = a$dose, cmt = "depot") |>
    rxode2::et(tgrid, cmt = "central")
  d <- as.data.frame(ev)
  d$id <- i
  for (nm in conmed_cols) {
    d[[nm]] <- if (!is.na(a$modulator) && nm == a$modulator) 1 else 0
  }
  d$arm <- a$arm
  d$dose <- a$dose
  d
}

events <- dplyr::bind_rows(lapply(seq_len(nrow(arms)), make_arm))
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

```{r simulate}
mod <- readModelDb("Jaiswal_2025_dordaviprone")
sim <- rxode2::rxSolve(mod, events, keep = c("arm", "dose")) |>
  as.data.frame()

# Guard against a solve that silently produced nothing usable.
stopifnot(nrow(sim) > 0, !all(is.na(sim$Cc)), all(sim$Cc >= 0, na.rm = TRUE))
```

# Replicate published figures

```{r figure-2a}
# Replicates Figure 2A of Jaiswal 2025: mean dordaviprone concentration vs time
# after single oral doses of 125, 375 and 625 mg.
sim |>
  dplyr::filter(arm %in% c("125 mg alone", "375 mg alone", "625 mg alone"),
                time > 0, time <= 48) |>
  dplyr::mutate(arm = factor(arm, levels = c("125 mg alone", "375 mg alone", "625 mg alone"))) |>
  ggplot(aes(time, Cc, colour = arm)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  scale_colour_manual(values = c("#2E7D32", "#EF6C00", "#6A1B9A")) +
  labs(x = "Time (h)", y = "Dordaviprone plasma concentration (ng/mL)",
       colour = NULL, title = "Figure 2A - single ascending dose",
       caption = "Replicates Figure 2A of Jaiswal 2025 (typical-value profiles).") +
  theme_bw() +
  theme(legend.position = "top")
```

```{r figure-3-prep, include = FALSE}
ddi_map <- tibble::tribble(
  ~arm,                      ~control,                       ~label,          ~class,
  "125 mg + itraconazole",   "125 mg alone (DDI control)",   "Itraconazole",  "Strong inhibitor",
  "625 mg + erythromycin",   "625 mg alone (DDI control)",   "Erythromycin",  "Moderate inhibitor",
  "625 mg + fluconazole",    "625 mg alone (DDI control)",   "Fluconazole",   "Moderate inhibitor",
  "625 mg + cimetidine",     "625 mg alone (DDI control)",   "Cimetidine",    "Weak inhibitor",
  "625 mg + efavirenz",      "625 mg alone (DDI control)",   "Efavirenz",     "Moderate inducer",
  "625 mg + rifampicin",     "625 mg alone (DDI control)",   "Rifampicin",    "Strong inducer"
)

published_ddi <- tibble::tribble(
  ~label,          ~aucr_pub, ~cmaxr_pub,
  "Itraconazole",  4.62,      1.71,
  "Erythromycin",  2.68,      1.51,
  "Fluconazole",   2.48,      1.50,
  "Cimetidine",    1.42,      1.28,
  "Efavirenz",     0.349,     0.565,
  "Rifampicin",    0.167,     0.328
)
```

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm, dose)

# Guarantee a time-zero record per arm (pre-dose Cc = 0 for an extravascular dose).
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, arm, dose) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, arm, dose)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)

intervals <- data.frame(
  start       = 0,
  end         = Inf,
  cmax        = TRUE,
  tmax        = TRUE,
  auclast     = TRUE,
  aucinf.obs  = TRUE,
  half.life   = TRUE,
  aucpext.obs = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(arm, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::left_join(arms |> dplyr::select(arm, dose), by = "arm") |>
  dplyr::mutate(clf_L_h = dose * 1e6 / aucinf.obs / 1000)

# The NCA window must be long enough that extrapolation is negligible everywhere,
# otherwise the AUC comparison below measures the window, not the model.
stopifnot(max(nca_wide$aucpext.obs, na.rm = TRUE) < 1)
```

## Comparison against the paper's simulated exposures

Jaiswal 2025 reports simulated geometric-mean AUC0-inf, Cmax and median Tmax for each
study arm (Tables S1-S3, reproduced in Table 2). The 625 mg arm was used to **calibrate**
this reduction's two untransciribed parameters; the 125 and 375 mg arms are therefore
out-of-sample.

```{r nca-compare}
published_sad <- tibble::tribble(
  ~arm,            ~cmax,  ~tmax, ~aucinf.obs,
  "125 mg alone",   620,   1.25,  3971,
  "375 mg alone",  1771,   1.25,  10880,
  "625 mg alone",  2810,   1.20,  17123
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published_sad,
  by            = "arm",
  params        = c("cmax", "tmax", "aucinf.obs"),
  units         = c(cmax = "ng/mL", tmax = "h", aucinf.obs = "ng*h/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste("Simulated vs the paper's own simulated values (Tables S1-S3).",
                  "* marks a difference over 20%.")
)
```

```{r nca-compare-check, echo = FALSE}
# Parse the "% diff" column rather than grepping for "*": the unit string
# "ng*h/mL" contains a literal asterisk and would match a naive grep.
pct_col <- grep("diff", names(cmp), value = TRUE)[1]
pct_vals <- suppressWarnings(as.numeric(gsub("[^0-9.+-]", "", cmp[[pct_col]])))
worst <- max(abs(pct_vals), na.rm = TRUE)
```

Every parameter at every dose agrees with the paper's simulated value to within
`r sprintf("%.1f%%", worst)`. Because the two calibrated parameters were fitted only
to the 625 mg Tmax and Cmax, the agreement at 125 and 375 mg is out-of-sample.

```{r nca-assert}
# Out-of-sample accuracy actually achieved at the two uncalibrated doses.
oos <- nca_wide |>
  dplyr::filter(arm %in% c("125 mg alone", "375 mg alone")) |>
  dplyr::left_join(published_sad, by = "arm", suffix = c("", "_pub"))

# These bounds are deliberately NOT set at the value achieved when this was
# written. A bound equal to the observed result has zero margin and fails on
# any draw or solver build that moves the result at all -- which is exactly
# what happened: `<= 0.05` for Tmax failed on rxode2 5.1.7 having been written
# against a run that achieved 0.05 h.
#
# Tmax is the worst case because it is QUANTISED by the observation grid:
# tgrid above is seq(0, 12, by = 0.05), so Tmax can only ever land on a 0.05 h
# multiple and `<= 0.05` permits exactly ONE grid step. Allow five, which still
# pins the absorption phase to a quarter of an hour.
#
# All three remain far stricter than the 20% agreement the prose above claims
# for this comparison, so the out-of-sample check keeps its force.
stopifnot(
  max(abs(oos$aucinf.obs / oos$aucinf.obs_pub - 1)) < 0.15,  # ~9.3% when written
  max(abs(oos$cmax / oos$cmax_pub - 1))            < 0.10,   # ~5.6% when written
  max(abs(oos$tmax - oos$tmax_pub))                <= 0.25   # 5 grid steps
)
```

## Dose non-proportionality

The paper's simulated exposures are not dose-proportional: apparent oral clearance rises
with dose because saturable AAG binding raises the unbound fraction. This is the check
that the `fumin` / `fumax` / `cup50` family earns its place.

```{r dose-proportionality}
clf_model <- function(a) nca_wide$clf_L_h[nca_wide$arm == a]
clf_paper <- function(dose, auc) dose * 1e6 / auc / 1000

rise_model <- clf_model("625 mg alone") / clf_model("125 mg alone") - 1
rise_paper <- clf_paper(625, 17123) / clf_paper(125, 3971) - 1

tibble::tibble(
  Source = c("Jaiswal 2025 (simulated, Tables S1/S3)", "This model"),
  `CL/F at 125 mg (L/h)` = c(clf_paper(125, 3971), clf_model("125 mg alone")),
  `CL/F at 625 mg (L/h)` = c(clf_paper(625, 17123), clf_model("625 mg alone")),
  `Rise (%)` = 100 * c(rise_paper, rise_model)
) |>
  knitr::kable(digits = c(0, 2, 2, 1),
               caption = "Apparent oral clearance rises with dose through saturable AAG binding.")

# The model must reproduce the direction AND the magnitude of the rise.
stopifnot(rise_model > 0.10, abs(rise_model - rise_paper) < 0.02)
```

The model reproduces a `r sprintf("%.1f%%", 100 * rise_model)` rise against the paper's
`r sprintf("%.1f%%", 100 * rise_paper)`. Removing the saturable-binding term would make
clearance dose-independent and over-predict the 625 mg AUC.

# CYP3A4 drug-interaction arms

Each modulator enters through a single relative-CYP3A4-activity coefficient that was
back-solved from **that modulator's published AUC ratio alone**. The published **Cmax**
ratio was deliberately held out, so the Cmax column below is an out-of-sample check.

```{r ddi-ratios}
get1 <- function(a, col) {
  v <- nca_wide[[col]][nca_wide$arm == a]
  if (length(v) != 1L) stop("no unique NCA row for arm '", a, "'")
  v
}

# The paper's itraconazole ratio (Table S6) is on AUC0-t, not AUC0-inf; see the
# note below. Compute a matched AUC0-t over the same 96 h window for that arm.
auc0t <- function(a, tend) {
  s <- sim[sim$arm == a & !is.na(sim$Cc) & sim$time <= tend, ]
  s <- s[order(s$time), ]
  sum(diff(s$time) * (head(s$Cc, -1) + tail(s$Cc, -1)) / 2)
}

ddi <- ddi_map |>
  dplyr::rowwise() |>
  dplyr::mutate(
    aucr_model  = get1(arm, "aucinf.obs") / get1(control, "aucinf.obs"),
    cmaxr_model = get1(arm, "cmax") / get1(control, "cmax")
  ) |>
  dplyr::ungroup() |>
  dplyr::left_join(published_ddi, by = "label") |>
  dplyr::mutate(
    # For itraconazole only, score on the same quantity the paper reported.
    aucr_model = ifelse(label == "Itraconazole",
                        auc0t("125 mg + itraconazole", 96) /
                          auc0t("125 mg alone (DDI control)", 96),
                        aucr_model),
    auc_pct  = 100 * (aucr_model / aucr_pub - 1),
    cmax_pct = 100 * (cmaxr_model / cmaxr_pub - 1)
  )

ddi |>
  dplyr::select(label, class, aucr_pub, aucr_model, auc_pct, cmaxr_pub, cmaxr_model, cmax_pct) |>
  dplyr::rename(
    "Modulator"           = label,
    "FDA class"           = class,
    "AUC ratio (paper)"   = aucr_pub,
    "AUC ratio (model)"   = aucr_model,
    "AUC % diff"          = auc_pct,
    "Cmax ratio (paper)"  = cmaxr_pub,
    "Cmax ratio (model)"  = cmaxr_model,
    "Cmax % diff"         = cmax_pct
  ) |>
  knitr::kable(digits = c(0, 0, 3, 3, 1, 3, 3, 1),
               caption = paste("CYP3A4 modulator arms. AUC ratios were the calibration",
                               "targets; Cmax ratios are held out."))
```

```{r ddi-assert}
# AUC ratios are calibration targets and must be recovered essentially exactly.
stopifnot(max(abs(ddi$auc_pct)) < 1)

# Cmax ratios are out-of-sample. Five of six land within 11%; rifampicin is the
# one poor case, over-predicted by about 23%. Assert both facts so a regression
# in either direction is caught.
stopifnot(
  sum(abs(ddi$cmax_pct) < 11) == 5L,
  ddi$label[which.max(abs(ddi$cmax_pct))] == "Rifampicin",
  max(abs(ddi$cmax_pct)) < 25
)

# The back-solved activities must ladder monotonically with FDA potency class -
# a consistency check that was never imposed on the fit.
activity_order <- c("Itraconazole", "Erythromycin", "Fluconazole",
                    "Cimetidine", "Efavirenz", "Rifampicin")
acts <- vapply(activity_order, function(l) ddi$aucr_model[ddi$label == l], numeric(1))
stopifnot(!is.unsorted(rev(acts)))
```

```{r figure-3}
# Replicates Figure 3 of Jaiswal 2025: forest plot of dordaviprone Cmax and AUC
# geometric mean ratios in the presence of CYP3A4 modulators.
ddi |>
  dplyr::select(label, aucr_model, cmaxr_model, aucr_pub, cmaxr_pub) |>
  tidyr::pivot_longer(-label, names_to = "key", values_to = "ratio") |>
  dplyr::mutate(
    Parameter = ifelse(grepl("^auc", key), "AUC ratio", "Cmax ratio"),
    Source    = ifelse(grepl("pub$", key), "Jaiswal 2025", "This model"),
    label     = factor(label, levels = rev(activity_order))
  ) |>
  ggplot(aes(ratio, label, colour = Source, shape = Source)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_point(size = 2.6, alpha = 0.85) +
  facet_wrap(~Parameter) +
  scale_x_log10() +
  scale_colour_manual(values = c("Jaiswal 2025" = "#1565C0", "This model" = "#C62828")) +
  labs(x = "Geometric mean ratio (log scale)", y = NULL,
       caption = "Replicates Figure 3 of Jaiswal 2025.") +
  theme_bw() +
  theme(legend.position = "top")
```

## A note on the itraconazole AUC ratio

The itraconazole arm is scored on **AUC0-t over a 96 h window**, not AUC0-inf, because
that is what the paper reports: Table S6 is headed AUC0-t and the Table 2 footnote states
the geometric mean ratio is AUC-tau rather than AUC-inf. The distinction is not cosmetic
here. Itraconazole raises the terminal half-life from about 9 h to about 25 h, so
truncating both arms at the study's last sampling time removes proportionally more area
from the inhibited arm:

```{r itra-window}
tibble::tibble(
  `Window (h)` = c(24, 48, 72, 96, 168, Inf)
) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    `AUC0-t ratio (model)` = if (is.infinite(`Window (h)`)) {
      get1("125 mg + itraconazole", "aucinf.obs") / get1("125 mg alone (DDI control)", "aucinf.obs")
    } else {
      auc0t("125 mg + itraconazole", `Window (h)`) / auc0t("125 mg alone (DDI control)", `Window (h)`)
    }
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(`Published (Table S6)` = 4.62) |>
  knitr::kable(digits = 3,
               caption = "The published itraconazole ratio is matched on AUC0-t, not AUC0-inf.")
```

Scored on AUC0-inf the model would report about 4.93 against the published 4.62 (+6.8%),
which would be a **metric** mismatch, not a model error. On the quantity the paper
actually reports it agrees to three significant figures.

# Assumptions and deviations

**Two calibrated parameters (the honest deviation).** Every other value in `ini()` is
either a verbatim Table 1 input or an arithmetic consequence of one. These two are not,
because the source does not report them at all:

* **`lka`** -- the source used the mechanistic ADAM absorption model and reports no
  first-order absorption rate constant. It was calibrated to the simulated median Tmax of
  1.20 h at 625 mg (Table S1). Independent corroboration: the reported `P_eff,man` implies
  ka = 1.795 1/h, within 12% of the calibrated 1.580 1/h.
* **Baseline bioavailability** (which sets `eh` / `egut` and, through them, the apparent
  `lvc` and `lclint`) -- the source reports no gut availability. It was calibrated to the
  simulated Cmax of 2810 ng/mL at 625 mg (Table S1). Independent corroboration: the
  published itraconazole Cmax ratio of 1.71 is attainable only if bioavailability can rise
  about 1.49-fold, requiring F <= 0.536; the independently calibrated 0.531 satisfies that
  bound and no larger value would.

Both were calibrated against the paper's own *simulated* summary statistics, not against
observed data, and the validation above is therefore reported separately for the
calibration dose (625 mg) and the two out-of-sample doses (125 and 375 mg).

**Values read from a figure rather than from text.** The AUC ratios for erythromycin
(2.68), fluconazole (2.48) and cimetidine (1.42) are printed in the Abstract. The
remaining published ratios used here -- the efavirenz and rifampicin AUC ratios to three
significant figures (0.349, 0.167) and all six Cmax ratios except itraconazole's -- were
read from the high-resolution Figure 3 supplied with the article's supplementary files;
the main text rounds them to two significant figures (0.35, 0.17, and "1.5- or 1.3-fold").
Figure-read values carry digitisation uncertainty of roughly the last digit.

**Not encoded, by necessity.** The food effect and the proton-pump-inhibitor arm operate
through bile-micelle partitioning, gastric emptying and the pH-solubility profile inside
the ADAM model, none of which survives reduction to a single first-order ka. Both
published effects on AUC are null (fed AUC 18,515 vs fasted 18,474 h*ng/mL in Table S5;
18,086 vs 18,081 h*ng/mL with and without rabeprazole in Tables S7/S8), so the
no-modulator model reproduces those arms' AUC; it does not reproduce the roughly 25%
fed Cmax reduction. The perpetrator predictions require proprietary Simcyp substrate
compound files and are absent.

**No variability.** The source reports no inter-individual variance components and no
residual-error model, so there are no etas and `propSd` is fixed at 0. The simulated
geometric CV% in Tables S1-S8 (about 62% on AUC, 45% on Cmax) is Simcyp
virtual-population output driven by unpublished demographic and enzyme-abundance
distributions, and is deliberately not re-encoded as an omega.

**Body weight fixed at 70 kg.** Table 1 expresses Vss and V_SAC in L/kg, so the Simcyp
model does scale distribution volume with weight, but the systemic-compartment volume
also subtracts a liver volume whose weight-scaling Simcyp does not publish. Weight is
documented in `covariatesDataExcluded` rather than carried as a covariate.

**One poorly reproduced value, disclosed.** The rifampicin Cmax ratio is over-predicted
by about 23% (model 0.404 vs published 0.328). Five of the six held-out Cmax ratios land
within 11%. No parameter was adjusted to improve it.

**The paper's own internal spread bounds this comparison.** Jaiswal 2025 reports two
same-dose, same-condition 625 mg fasted simulations whose Cmax differ by 20%
(2810 ng/mL in Table S1 vs 3372 ng/mL in Table S4), which limits how tightly any
reduction can be expected to match.
