---
title: "Oxcarbazepine (Jang 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Oxcarbazepine (Jang 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r readModelDb("Jang_2021_oxcarbazepine")()$reference`
* Description: `r readModelDb("Jang_2021_oxcarbazepine")()$description`
* Article: <https://doi.org/10.1038/s41598-021-85920-0> (open access)
* Supplementary material: the NONMEM control stream of the final model
  (`41598_2021_85920_MOESM2_ESM.docx`) and the goodness-of-fit figure plus
  co-medication table (`41598_2021_85920_MOESM1_ESM.docx`), both linked from
  the article page.

Jang 2021 built the population PK model to convert sparse spot concentrations
of the active monohydroxy derivative (MHD, licarbazepine) into individual
exposure metrics -- a steady-state trough and `AUC = Dose / (CL/F)` -- and then
related those metrics to dose-related adverse events (dizziness, somnolence,
headache, diplopia). The PK model is the part extracted here. Oxcarbazepine
itself is not modelled: its exposure is about 1/15 of MHD's, so each
oxcarbazepine dose (mg) enters the depot and the central compartment holds MHD
in oxcarbazepine-mg equivalents (the molar masses are 252.3 and 254.3 g/mol, so
the two scales agree to within 1%).

# Population

The PK dataset combined two sources (Jang 2021 Methods):

* **Study 1**, the Epilepsy Registry Cohort of Seoul National University
  Hospital (February 2011 to January 2014): single spot samples from adults on
  an unchanged oxcarbazepine dose for at least one month, assumed to be at
  steady state. Of 618 patients with 711 samples, 447 had complete body-weight,
  last-dose and sampling-time records. 438 took oxcarbazepine twice daily, 5
  once daily and 4 three times daily.
* **Study 2**, a single-dose oral-loading study in 40 patients with epilepsy
  (30 mg/kg; Kim DW et al. Epilepsia 2012;53:e9-12), sampled at 2, 4, 6, 8, 12,
  14, 16 and 24 h.

In total 748 MHD concentrations from 487 patients were modelled. Table 1 and the
Results describe the 447 registry patients: 59.1% male, mean age 39.2 years
(range 16-80), mean body weight 65.8 kg (range 39-116), mean daily dose 999 mg
(range 150-2100 mg/day), 65.3% on at least one other antiseizure medication
(most often levetiracetam, 55.5%, and topiramate, 32.9%) and 5.1% on an
enzyme-inducing one. All patients were Korean.

```{r population}
str(readModelDb("Jang_2021_oxcarbazepine")()$population)
```

# Source trace

Every `ini()` entry in `inst/modeldb/specificDrugs/Jang_2021_oxcarbazepine.R`
carries an in-file comment naming its source. The final estimates come from Jang
2021 Table 2; the model structure comes from the supplementary NONMEM control
stream, whose `$THETA` / `$OMEGA` records are initial estimates and are not used
as values.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| One compartment, first-order absorption (`ADVAN2 TRANS2`) | n/a | Methods, Population PK model development; supplementary code `$SUBROUTINE` |
| `CL/F = theta1 * exp(eta1) * (WT/66)^theta2` | n/a | Table 2 structural block; supplementary code `CL = TVCL*EXP(ETA(1))*(BW/66)**THETA(4)` |
| `V/F = theta3 * (WT/66)^theta4` | n/a | Table 2 structural block; supplementary code `V = TVV*EXP(ETA(2))*(BW/66)**THETA(5)` with `$OMEGA 0 FIX` for `ETA(2)` |
| `ka = theta * exp(eta3)` | n/a | supplementary code `KA = TVKA*EXP(ETA(3))` |
| `lcl` (CL/F at 66 kg) | 1.65 L/h (1.8% RSE) | Table 2 theta1; bootstrap 1.65 (1.59-1.70) |
| `e_wt_cl` | 0.67 (14.2% RSE) | Table 2 theta2; bootstrap 0.66 (0.47-0.85) |
| `lvc` (V/F at 66 kg) | 59.0 L (4.5% RSE) | Table 2 theta3 (unit printed as L/h, a typo for L); bootstrap 59.2 (54.2-64.8) |
| `e_wt_vc` | 0.96 (18.1% RSE) | Table 2 theta4; bootstrap 0.95 (0.59-1.3) |
| `lka` | 0.34 1/h (9.7% RSE) | Table 2 ka; bootstrap 0.34 (0.28-0.41) |
| `etalcl` | 29.2 %CV -> `omega^2 = log(1 + 0.292^2) = 0.08183` | Table 2 'IIV CL/F' |
| `etalka` | 41.5 %CV -> `omega^2 = log(1 + 0.415^2) = 0.15952` | Table 2 second IIV row (labelled 'IIV V2/F'); assigned to ka per the control stream, see below |
| `propSd` | 0.13 | Table 2 'Proportional error (SD)'; supplementary code `W = SQRT(THETA(6)**2 + THETA(7)**2*IPRED**2)` with `$SIGMA 1 FIX` |
| Additive error `THETA(6)` | 0.0001 ng/mL, fixed | supplementary code; omitted (see Assumptions) |

# Which parameter carries the second IIV?

Table 2 labels its second inter-individual variability "IIV V2/F" (41.5 %CV),
but the supplementary control stream -- captioned "NONMEM code for the final
model" -- fixes the variance of `ETA(2)` on V to zero and estimates `ETA(3)` on
KA. Both sources list exactly two estimated variances, so the question is only
which parameter the 41.5 %CV belongs to.

Figure 1 (right panel, Study 2) settles it. For a single mg/kg dose the
body-weight terms nearly cancel, so the width of the simulated 5th-95th
percentile band around the peak is set by the IIV: 41.5 %CV on V/F scales the
whole curve and gives a wide band, while 41.5 %CV on ka mostly shifts the rise.
The maintainers digitised the simulated percentile bands of Figure 1 (centre of
each shaded band) and compared them with both readings, using the closed-form
one-compartment solution and base-R draws so the comparison is identical on
every machine.

```{r iiv-location}
# Figure 1, Study 2: centres of the simulated percentile bands digitised from
# the published raster (time in h, concentration in mg/L).
fig1 <- tibble::tribble(
  ~time, ~p05, ~p50, ~p95,
  6,     18.1, 24.8, 32.8,
  12,    18.3, 24.2, 30.0,
  20,    12.9, 19.3, 26.0
)

closed_form <- function(time, dose, cl, v, ka) {
  k <- cl / v
  dose / v * ka / (ka - k) * (exp(-k * time) - exp(-ka * time))
}

set.seed(20210318)
n_draw <- 2000
wt <- numeric(0)
while (length(wt) < n_draw) {
  x <- rnorm(n_draw, 65.8, 12.5)
  wt <- c(wt, x[x >= 39 & x <= 116])
}
wt <- wt[seq_len(n_draw)]
eta_cl <- rnorm(n_draw, 0, sqrt(log(1 + 0.292^2)))
eta_2 <- rnorm(n_draw, 0, sqrt(log(1 + 0.415^2)))
eps <- rnorm(n_draw, 0, 0.13)

band <- function(eta_v, eta_ka) {
  bind_rows(lapply(fig1$time, function(tm) {
    conc <- closed_form(
      tm,
      dose = 30 * wt,
      cl = 1.65 * exp(eta_cl) * (wt / 66)^0.67,
      v = 59.0 * exp(eta_v) * (wt / 66)^0.96,
      ka = 0.34 * exp(eta_ka)
    ) * (1 + eps)
    tibble(
      time = tm,
      p05 = quantile(conc, 0.05, names = FALSE),
      p50 = quantile(conc, 0.50, names = FALSE),
      p95 = quantile(conc, 0.95, names = FALSE)
    )
  }))
}

hyp <- bind_rows(
  band(eta_v = 0, eta_ka = eta_2) |> mutate(reading = "IIV on ka (control stream)"),
  band(eta_v = eta_2, eta_ka = 0) |> mutate(reading = "IIV on V/F (Table 2 label)")
) |>
  left_join(fig1, by = "time", suffix = c("", "_fig"))

hyp |>
  transmute(
    Reading = reading, `Time (h)` = time,
    `5th` = round(p05, 1), `5th, Fig. 1` = p05_fig,
    `50th` = round(p50, 1), `50th, Fig. 1` = p50_fig,
    `95th` = round(p95, 1), `95th, Fig. 1` = p95_fig
  ) |>
  knitr::kable(caption = paste(
    "Simulated percentiles (mg/L) after a single 30 mg/kg dose under the two",
    "readings of the second IIV, against the Figure 1 simulated bands."
  ))

ka_reading <- filter(hyp, reading == "IIV on ka (control stream)")
v_reading <- filter(hyp, reading == "IIV on V/F (Table 2 label)")
# The control-stream reading reproduces every Figure 1 band within 15%.
stopifnot(
  all(abs(ka_reading$p05 / ka_reading$p05_fig - 1) < 0.15),
  all(abs(ka_reading$p50 / ka_reading$p50_fig - 1) < 0.15),
  all(abs(ka_reading$p95 / ka_reading$p95_fig - 1) < 0.15)
)
# The Table 2-label reading puts the 6 h 95th percentile far above the
# published band (about 47 vs 33 mg/L).
stopifnot(v_reading$p95[v_reading$time == 6] > 1.25 * 32.8)
```

The control-stream reading reproduces the published bands; the other reading
puts the 6 h 95th percentile about 45% above the figure. The model therefore
places the 41.5 %CV on ka, and the "V2/F" label in Table 2 is treated as a
typographical error.

# Typical-value identity

At the 66 kg reference weight, with random effects removed, the packaged model
must reproduce the closed-form one-compartment solution with the Table 2
estimates.

```{r identity}
mod <- readModelDb("Jang_2021_oxcarbazepine")
mod_typ <- rxode2::zeroRe(mod)

typ_ev <- bind_rows(
  tibble(id = 1L, time = 0, evid = 1L, amt = 30 * 66, cmt = "depot"),
  tibble(id = 1L, time = c(0, 1, 2, 4, 6, 8, 12, 16, 24, 48), evid = 0L,
         amt = 0, cmt = "central")
) |>
  mutate(WT = 66)

typ <- rxode2::rxSolve(mod_typ, typ_ev, returnType = "data.frame",
                       atol = 1e-10, rtol = 1e-10)
expected <- closed_form(typ$time, dose = 30 * 66, cl = 1.65, v = 59.0, ka = 0.34)
stopifnot(
  all(abs(typ$cl - 1.65) < 1e-10),
  all(abs(typ$vc - 59.0) < 1e-10),
  all(abs(typ$Cc - expected) <= 1e-6 * max(expected))
)
knitr::kable(
  tibble(`Time (h)` = typ$time, `Model Cc (mg/L)` = signif(typ$Cc, 5),
         `Closed form (mg/L)` = signif(expected, 5)),
  caption = "Typical 66 kg patient, single 30 mg/kg dose."
)
```

# Virtual cohort

Two cohorts of 200 virtual patients each, with body weight drawn from the
registry distribution (mean 65.8 kg, SD 12.5 kg, redrawn outside the observed
39-116 kg range):

* **Single dose** -- 30 mg/kg once, as in Study 2.
* **Steady state** -- twice-daily dosing at a total daily dose drawn from the
  registry distribution (mean 999 mg/day, SD 338 mg/day, redrawn outside the
  observed 150-2100 mg/day range), as in Study 1.

```{r cohort}
draw_in_range <- function(n, mean, sd, lower, upper) {
  out <- numeric(0)
  while (length(out) < n) {
    x <- rnorm(n, mean, sd)
    out <- c(out, x[x >= lower & x <= upper])
  }
  out[seq_len(n)]
}

set.seed(6370)
n_sub <- 200
sd_subj <- tibble(id = seq_len(n_sub),
                  WT = draw_in_range(n_sub, 65.8, 12.5, 39, 116),
                  treatment = "30 mg/kg single dose")
ss_subj <- tibble(id = n_sub + seq_len(n_sub),
                  WT = draw_in_range(n_sub, 65.8, 12.5, 39, 116),
                  daily_dose = draw_in_range(n_sub, 999, 338, 150, 2100),
                  treatment = "Twice daily, steady state")

obs_sd <- c(0, 0.5, 1, 1.5, 2, 3, 4, 5, 6, 7, 8, 10, 12, 14, 16, 18, 20, 22, 24)
sd_ev <- bind_rows(
  sd_subj |> mutate(time = 0, evid = 1L, amt = 30 * WT, ii = 0, ss = 0L,
                    cmt = "depot"),
  sd_subj |> tidyr::crossing(time = obs_sd) |>
    mutate(evid = 0L, amt = 0, ii = 0, ss = 0L, cmt = "central")
)
# ss = 1 at time 0 puts the subject at steady state; addl = 1 gives the
# second dose of the 24 h window.
ss_ev <- bind_rows(
  ss_subj |> mutate(time = 0, evid = 1L, amt = daily_dose / 2, ii = 12,
                    ss = 1L, addl = 1L, cmt = "depot"),
  ss_subj |> tidyr::crossing(time = seq(0, 24, by = 0.5)) |>
    mutate(evid = 0L, amt = 0, ii = 0, ss = 0L, addl = 0L, cmt = "central")
)
events <- bind_rows(sd_ev, ss_ev) |>
  mutate(addl = coalesce(addl, 0L)) |>
  select(id, time, evid, amt, ii, ss, addl, cmt, WT, treatment) |>
  arrange(id, time, desc(evid))
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
rxode2::rxSetSeed(6370)
# steady-state searches for long-half-life subjects exceed the default step budget
sim <- rxode2::rxSolve(mod, events, keep = "treatment",
                       returnType = "data.frame", maxsteps = 1e6) |>
  filter(!is.na(Cc))
```

## Replicating Figure 1, Study 2

```{r fig1-study2, fig.cap = "Replicates the right panel (Study 2) of Figure 1 of Jang 2021: simulated 5th, 50th and 95th percentiles after a single 30 mg/kg dose (shaded band and line), with the centres of the published simulated percentile bands as points. Residual error is included, as in the published VPC."}
vpc <- sim |>
  filter(treatment == "30 mg/kg single dose") |>
  group_by(time) |>
  summarise(p05 = quantile(sim, 0.05), p50 = quantile(sim, 0.50),
            p95 = quantile(sim, 0.95), .groups = "drop")

fig1_long <- fig1 |>
  pivot_longer(-time, names_to = "percentile", values_to = "conc")

ggplot(vpc, aes(time)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = p50), colour = "steelblue4", linewidth = 1) +
  geom_point(data = fig1_long, aes(y = conc, shape = percentile), size = 2.5) +
  labs(x = "Time after dose (h)", y = "MHD concentration (mg/L)",
       shape = "Figure 1 band") +
  theme_bw()

vpc_chk <- vpc |>
  filter(time %in% fig1$time) |>
  left_join(fig1, by = "time", suffix = c("", "_fig"))
# Cohort percentiles differ between rxode2 builds; the median must sit near
# the published band and the tails within a wider envelope.
stopifnot(
  all(abs(vpc_chk$p50 / vpc_chk$p50_fig - 1) < 0.15),
  all(abs(vpc_chk$p05 / vpc_chk$p05_fig - 1) < 0.30),
  all(abs(vpc_chk$p95 / vpc_chk$p95_fig - 1) < 0.30)
)
```

The Study 1 panel of Figure 1 is not replicated: it pools spot samples across
the registry's individual doses, regimens and last-dose times, which are not
published.

# PKNCA validation

Non-compartmental analysis on the model-predicted concentrations (no residual
error): the single-dose profile over 0-24 h and one steady-state day
(0-24 h, two doses) of twice-daily dosing.

```{r pknca}
nca_conc <- sim |>
  select(id, time, Cc = ipredSim, treatment) |>
  filter(!is.na(Cc))
nca_dose <- events |>
  filter(evid == 1L) |>
  select(id, time, amt, treatment)
nca_dose <- bind_rows(
  nca_dose,
  nca_dose |> filter(treatment == "Twice daily, steady state") |>
    mutate(time = 12)
)

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ time | treatment + id,
                             concu = "mg/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(nca_dose, amt ~ time | treatment + id,
                             doseu = "mg")
intervals <- data.frame(start = 0, end = 24, cmax = TRUE, tmax = TRUE,
                        cmin = TRUE, auclast = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
summary(nca_res)
```

## Steady-state AUC over 24 h equals daily dose / (CL/F)

Jang 2021 computed each patient's AUC as `Dose / (CL/F)` with the daily dose.
For the steady-state cohort the NCA AUC over one 24 h day must reproduce that
identity per subject, up to trapezoidal error on the 0.5 h grid.

```{r auc-identity}
ind_cl <- sim |>
  filter(treatment == "Twice daily, steady state") |>
  distinct(id, cl)
auc_chk <- as.data.frame(nca_res) |>
  filter(treatment == "Twice daily, steady state", PPTESTCD == "auclast") |>
  select(id, auc24 = PPORRES) |>
  left_join(ss_subj |> select(id, daily_dose), by = "id") |>
  left_join(ind_cl, by = "id") |>
  mutate(pct_diff = 100 * (auc24 / (daily_dose / cl) - 1))
stopifnot(
  nrow(auc_chk) == n_sub,
  abs(median(auc_chk$pct_diff)) < 1,
  quantile(abs(auc_chk$pct_diff), 0.9) < 2
)
summary(auc_chk$pct_diff)
```

## Comparison with Jang 2021 Table 1

Table 1 reports, for the 447 registry patients, the mean individual MHD AUC
(`daily dose / (CL/F)` from the post hoc CL/F) and the mean model-derived
trough. Both are means, so the simulated steady-state cohort is summarised by
its mean here rather than the median.

```{r compare-table1}
ss_nca <- as.data.frame(nca_res) |>
  filter(treatment == "Twice daily, steady state",
         PPTESTCD %in% c("auclast", "cmin")) |>
  group_by(PPTESTCD) |>
  summarise(PPORRES = mean(PPORRES), .groups = "drop")

published <- data.frame(auclast = 653.8, cmin = 12.75)
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = ss_nca,
  reference = published,
  units = c(auclast = "h*mg/L", cmin = "mg/L"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = paste(
  "Mean steady-state exposure over one 24 h day of twice-daily dosing:",
  "simulated cohort vs Jang 2021 Table 1 (447 registry patients).",
  "AUC0-last spans 0-24 h, i.e. the daily AUC."
))

auc_sim <- ss_nca$PPORRES[ss_nca$PPTESTCD == "auclast"]
cmin_sim <- ss_nca$PPORRES[ss_nca$PPTESTCD == "cmin"]
stopifnot(abs(auc_sim / 653.8 - 1) < 0.15)
# The trough discrepancy described below is large, not cohort noise.
stopifnot(cmin_sim > 1.4 * 12.75)

# Typical 66 kg patient at steady state on 500 mg twice daily: ratio of the
# trough to the daily-average concentration.
typ_ss <- rxode2::rxSolve(
  mod_typ,
  bind_rows(
    tibble(id = 1L, time = 0, evid = 1L, amt = 500, ii = 12, ss = 1L,
           cmt = "depot"),
    tibble(id = 1L, time = 12, evid = 0L, amt = 0, ii = 0, ss = 0L,
           cmt = "central")
  ) |> mutate(WT = 66),
  returnType = "data.frame", maxsteps = 1e6
)
trough_ratio <- typ_ss$Cc / (1000 / 24 / 1.65)
stopifnot(trough_ratio > 0.85, trough_ratio < 0.95)
round(c(mean_trough_sim = cmin_sim, typical_trough_to_average = trough_ratio), 2)
```

The daily AUC agrees with Table 1. The trough does not: the simulated mean
twice-daily trough is well above the 12.75 mg/L in Table 1. The two Table 1
columns are not consistent with each other under the published model either --
with CL/F = 1.65 L/h and V/F = 59 L the half-life is about 25 h, so for the
typical patient a twice-daily trough is about 90% of the daily-average
concentration (`AUC / 24` = 27 mg/L for the Table 1 mean AUC). Jang 2021
describes the trough only as "estimated using a simulation method", without the
regimen or timing used, so the discrepancy cannot be traced from the article.
It is reported here and was not used to change any parameter. The mean observed
spot MHD concentration in Table 1 (15.8 mg/L) is likewise below the
daily-average concentration the model implies for the mean dose.

# Adverse-event analysis (not extracted)

Jang 2021 related the individual MHD trough and AUC to dose-related adverse
events with ROC cut-offs (dose-related adverse events: trough 12.27 mg/L, AUC
698.5 mg h/L; dizziness alone: trough 19.15 mg/L, AUC 940.5 mg h/L) and a
multivariable logistic regression adjusted for age, sex and GFR (Table 3; for
dizziness, OR 1.079 per mg/L of trough and 1.002 per mg h/L of AUC). The
regression intercepts are not reported, so no probability model can be
reconstructed; the cut-offs are thresholds, not a model. Neither is packaged.

# Assumptions and deviations

* **Second IIV on ka, not V/F.** Table 2 labels the 41.5 %CV "IIV V2/F"; the
  final-model control stream fixes the V variance to zero and estimates the ka
  variance. The control stream is followed, confirmed against the Figure 1
  simulated bands above.
* **%CV to variance.** Table 2 gives IIV as %CV of an exponential model; the
  variances use `omega^2 = log(1 + CV^2)`. With `omega^2 = CV^2` instead, the
  variances would be 0.0853 and 0.1722, a difference that does not affect any
  check above.
* **Negligible additive error omitted.** The control stream's error model is
  `W = sqrt(THETA(6)^2 + THETA(7)^2 * IPRED^2)` with `THETA(6)` fixed at 0.0001
  ng/mL; the paper describes the error model as proportional, and the fixed
  additive term (1e-7 mg/L) is omitted.
* **Concentration units.** The analysis dataset was in ng/mL (`SC = V/1000`
  with doses in mg). The model reports mg/L; Table 2 prints the proportional
  error with a unit of ng/mL, which does not apply to a fractional SD.
* **Table 2 unit typo.** The V/F typical value is printed with unit "L h-1";
  it is a volume in L, as the equation line above it states.
* **Body weight of the Study 2 patients.** Not reported; the virtual
  single-dose cohort uses the registry weight distribution.
* **Screened, not retained.** Age, sex and enzyme-inducing co-medication (an
  estimated 7% increase in CL/F) were tested and not retained; they are listed
  in `covariatesDataExcluded`.
* **Trough and observed levels.** See the comparison with Table 1 above: the
  published model reproduces the Table 1 AUC but not the Table 1 trough, and
  the article does not describe the trough simulation in enough detail to find
  the cause.
