---
title: "Zaltoprofen (Jang 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Zaltoprofen (Jang 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Jang_2023_zaltoprofen"))$reference`
* Article: <https://doi.org/10.3390/ph16020161>
* Supplement (Tables S1-S6, the sole source of every parameter value):
  <https://www.mdpi.com/article/10.3390/ph16020161/s1>

# READ THIS FIRST: the published parameter set is internally inconsistent

**The parameter values packaged here are exactly as printed in the source
paper's supplementary Table S3, and they do not reproduce the source paper's
own Figures 1-2 or its quoted steady-state concentrations.** At the published
`tv CL/F = 43.70 L/h`, an 80 mg oral dose gives `AUC(0-inf) = 1.83 ug*h/mL`
and `Tmax = 0.21 h`, whereas the observed means the same paper plots in
Figure 1A peak near 3.6 ug/mL at about 1 h with `AUC(0-24)` around
12.5 ug*h/mL, and the paper's own text quotes a steady-state mean of
0.43 ug/mL where these parameters give 0.076 ug/mL. The discrepancy is
roughly 6-7 fold in exposure and roughly 5 fold in Tmax.

The values are shipped verbatim anyway, under the standing rule that printed
values have authority and that figure-fitting never overrides print
(operator ruling, 2026-09-02). Refitting to the figure would produce an
original model rather than this paper's model.

Consequently **this vignette validates self-consistency, not agreement with
the paper's figures**: it checks that the packaged encoding faithfully
reproduces Equations (1)-(5) and Table S3, that the solved ODE system matches
the closed-form two-compartment oral solution, and that the covariate
sub-model reproduces the ratios the paper itself publishes. The
[Errata](#errata-the-table-s3-versus-figures-conflict) section at the end
carries the full quantitative demonstration of the conflict.

Do not use this model to predict absolute zaltoprofen exposure without
reading that section first.

# Population

The model was fitted to the pharmacokinetic arm of a bioequivalence study in
**26 healthy Korean male adults** given a **single 80 mg oral dose** of
zaltoprofen (Methods 4.2, citing the authors' earlier report). Baseline
characteristics (supplementary Table S5) were age 23.19 +/- 2.26 years,
weight 64.73 +/- 8.08 kg, height 172.64 +/- 5.95 cm, serum albumin
4.92 +/- 0.18 g/dL, serum creatinine 0.98 +/- 0.09 mg/dL, Cockcroft-Gault
creatinine clearance 107.53 +/- 17.28 mL/min and MDRD glomerular filtration
rate 101.34 +/- 11.53 mL/min. Every biochemical parameter sat within normal
limits, so the model carries no information about renal or hepatic
impairment.

CYP2C9 was genotyped by PCR-RFLP; only the `*1` and `*3` alleles were
detected, so subjects were either `*1/*1` or `*1/*3`. Genotype was screened
as a covariate and rejected (see below).

The model was fitted in Phoenix NLME 8.3 by first-order conditional
estimation with extended least squares and eta-epsilon interaction, and
qualified with goodness-of-fit plots, a 1000-replicate non-parametric
bootstrap (Table S4), a 100-simulation VPC (Figures S7-S8) and NPDE
(Figure S6). Published mean profiles from three earlier studies (Table S6:
Kang 2006, Lee 2006, Li 2011) were used for external validation only and did
not inform the fit.

The same information is available programmatically via
`readModelDb("Jang_2023_zaltoprofen")()$population`.

# Structural model

Two-compartment disposition with first-order absorption and **no** lag time.
Table S1 selected two compartments over one (`d-2LL = -2210.5`) and over
three (`d-2LL = -0.13`, no gain), rejected a lag time (`d-2LL = +70.58`),
selected a proportional residual error over additive (`d-2LL = -474.97`), and
removed the etas on `V/F` and `CL2/F` (step 02-04-06).

The paper's Equations (1)-(5) are, verbatim:

```
V/F   = tv V/F                                                                  (1)
CL/F  = tv CL/F * (CrCL/mCrCL)^dCL/FdCrCL * (Albumin/mAlbumin)^dCL/FdAlbumin
                * exp(etaCL/F)                                                  (2)
V2/F  = tv V2/F * exp(etaV2/F)                                                  (3)
CL2/F = tv CL2/F                                                                (4)
Ka    = tv Ka * exp(etaKa)                                                       (5)
```

with `mCrCL = 104.38 mL/min` and `mAlbumin = 4.90 g/dL` (the cohort medians,
stated in Results 2.4 and repeated in the Figure 2-5 captions). The text
immediately after Eq. (5) states that `V` and `CL2` are "volume and clearance
in the central and peripheral compartments", matching Phoenix NLME's native
`V / Cl / V2 / Cl2` two-compartment parameterisation.

# Source trace

Every value below is from supplementary Table S3, "Estimate" column. The
in-file comments in `inst/modeldb/specificDrugs/Jang_2023_zaltoprofen.R`
carry the same trace per parameter.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lka` (Ka) | 1.73 1/h | Table S3, `tv Ka` (SE 0.15, RSE 8.56%) |
| `lvc` (V/F) | 4.88 L | Table S3, `tv V/F` (SE 1.28, RSE 26.32%) |
| `lcl` (CL/F) | 43.70 L/h | Table S3, `tv CL/F` (SE 2.12, RSE 4.85%) |
| `lvp` (V2/F) | 40.56 L | Table S3, `tv V2/F` (SE 7.02, RSE 17.32%) |
| `lq` (CL2/F) | 5.61 L/h | Table S3, `tv CL2/F` (SE 0.63, RSE 11.16%) |
| `e_crcl_cl` | 0.48 | Table S3, `dCL/FdCrCL` (SE 0.17, RSE 35.09%) |
| `e_alb_cl` | -1.83 | Table S3, `dCL/FdAlbumin` (SE 0.64, RSE 35.18%) |
| `etalka` | 0.16508 | Table S3, `omega^2 Ka` = 0.17; IIV 40.63% (see IIV note) |
| `etalcl` | 0.009604 | Table S3, `omega^2 CL/F` = 0.01; IIV 9.80% |
| `etalvp` | 0.23571 | Table S3, `omega^2 V2/F` = 0.24; IIV 48.55% |
| `propSd` | 0.38 | Table S3, `epsilon` (SE 0.02, RSE 6.03%) |
| CrCL centring 104.38 mL/min | n/a | Results 2.4 and Figure 2-5 captions |
| Albumin centring 4.90 g/dL | n/a | Results 2.4 and Figure 2-5 captions |
| `cl <- ... * (CRCL/104.38)^... * (alb/4.90)^...` | n/a | Eq. (2), page 3 |
| `vc <- exp(lvc)` (no eta) | n/a | Eq. (1), page 3; Table S1 step 02-04-06 |
| `vp <- exp(lvp + etalvp)` | n/a | Eq. (3), page 3 |
| `q <- exp(lq)` (no eta) | n/a | Eq. (4), page 3; Table S1 step 02-04-06 |
| `ka <- exp(lka + etalka)` | n/a | Eq. (5), page 3 |
| First-order absorption, no lag | n/a | Table S1, model 02 selected over 02-01 |
| Proportional residual error | n/a | Table S1, model 02-04 selected |

```{r model}
mod <- readModelDb("Jang_2023_zaltoprofen")
ui  <- rxode2::rxode(mod)
```

## IIV: which column of Table S3 is authoritative

Table S3 reports each between-subject term twice: an `omega^2` column rounded
to two decimals, and an `IIV (%)` column with four significant figures. The
two are related by `omega = IIV(%) / 100` (the Phoenix / NONMEM approximate-CV
convention), which reproduces every rounded `omega^2` in the table. The
alternative log-normal reading `omega = sqrt(log(1 + CV^2))` does not. The
model therefore takes the higher-precision values back-calculated from the
`IIV (%)` column.

```{r iiv-precision-check}
iiv_pct    <- c(Ka = 40.63, `CL/F` = 9.80, `V2/F` = 48.55)
omega2_tab <- c(Ka = 0.17,  `CL/F` = 0.01, `V2/F` = 0.24)

iiv_chk <- tibble(
  Parameter          = names(iiv_pct),
  `IIV (%)`          = iiv_pct,
  `omega^2 = (IIV/100)^2` = (iiv_pct / 100)^2,
  `rounds to`        = round((iiv_pct / 100)^2, 2),
  `Table S3 omega^2` = omega2_tab,
  `log-normal alternative` = round(log(1 + (iiv_pct / 100)^2), 2)
)
knitr::kable(iiv_chk, digits = 5,
             caption = "The approximate-CV reading reproduces all three printed variances; the log-normal reading does not.")

# Gate: exact rounding identity for the adopted reading, and a demonstration
# that the rejected reading genuinely disagrees (so this check can go red).
stopifnot(
  all(round((iiv_pct / 100)^2, 2) == omega2_tab),
  sum(round(log(1 + (iiv_pct / 100)^2), 2) != omega2_tab) >= 2L
)

# And the variances actually packaged in ini() are those values.
omega_packaged <- diag(ui$omega)
stopifnot(all(abs(sort(omega_packaged) - sort((iiv_pct / 100)^2)) < 5e-6))
```

# Self-consistency validation

The paper's figures cannot be used as the validation target (see the warning
above), so the packaged encoding is checked against quantities that follow
from Equations (1)-(5) and Table S3 alone. These are deterministic identities,
so the tolerances below are tight on purpose: each is many orders of magnitude
above the numerical error actually achieved, and each still goes red on a
mis-transcribed value.

```{r closed-form}
# Closed-form solution of a two-compartment model with first-order input,
# used as an independent implementation to check the packaged ODE system.
cf_2cmt_oral <- function(t, dose, ka, vc, cl, vp, q) {
  k10 <- cl / vc
  k12 <- q / vc
  k21 <- q / vp
  b     <- k10 + k12 + k21
  disc  <- sqrt(b^2 - 4 * k10 * k21)
  alpha <- (b + disc) / 2
  beta  <- (b - disc) / 2
  A <- (dose / vc) * ka * (k21 - alpha) / ((ka - alpha) * (beta - alpha))
  B <- (dose / vc) * ka * (k21 - beta)  / ((ka - beta)  * (alpha - beta))
  A * exp(-alpha * t) + B * exp(-beta * t) - (A + B) * exp(-ka * t)
}

# Terminal (beta) disposition rate constant of the published parameter set.
beta_rate <- local({
  k10 <- 43.70 / 4.88
  k12 <- 5.61 / 4.88
  k21 <- 5.61 / 40.56
  b   <- k10 + k12 + k21
  (b - sqrt(b^2 - 4 * k10 * k21)) / 2
})
t_half_beta <- log(2) / beta_rate
```

```{r typical-solve}
# Typical-value solve at the covariate reference subject (both ratio terms
# equal 1 there), on a grid fine enough to resolve a Tmax near 0.2 h.
grid_fine <- sort(unique(c(seq(0, 2, by = 0.002), seq(2, 24, by = 0.01))))

ev_tv <- rxode2::et(amt = 80, cmt = "depot") |>
  rxode2::et(grid_fine, cmt = "central") |>
  as.data.frame()
ev_tv$CRCL <- 104.38
ev_tv$ALB  <- 49.0          # 4.90 g/dL in canonical SI g/L

mod_tv <- rxode2::zeroRe(mod)
sim_tv <- rxode2::rxSolve(mod_tv, ev_tv, returnType = "data.frame") |>
  dplyr::filter(!is.na(Cc))

stopifnot(all(sim_tv$Cc >= 0))
```

## Check 1 - the ODE system matches the closed-form solution

```{r check-closed-form}
cf <- cf_2cmt_oral(sim_tv$time, dose = 80, ka = 1.73, vc = 4.88,
                   cl = 43.70, vp = 40.56, q = 5.61)
keep <- sim_tv$time > 0
rel_cf <- max(abs(cf[keep] - sim_tv$Cc[keep]) / cf[keep])
cat(sprintf("max relative difference, closed form vs ODE solve: %.3e\n", rel_cf))

# Achieved ~6e-14. The bound leaves six orders of magnitude of headroom and
# still fails instantly on any mis-transcribed structural value.
stopifnot(rel_cf < 1e-8)
```

## Check 2 - AUC(0-inf) equals Dose / (CL/F)

`AUC(0-inf) = Dose / CL` holds for any linear model with first-order input,
independent of the compartment structure, so it is the sharpest available
test that the clearance actually encoded is the clearance printed in
Table S3.

```{r check-auc}
auc_trap <- sum(diff(sim_tv$time) *
                  (head(sim_tv$Cc, -1) + tail(sim_tv$Cc, -1)) / 2)
auc_inf  <- auc_trap + tail(sim_tv$Cc, 1) / beta_rate
auc_expected <- 80 / 43.70

cat(sprintf("AUC(0-inf) solved   = %.6f ug*h/mL\n", auc_inf))
cat(sprintf("Dose / (CL/F)       = %.6f ug*h/mL\n", auc_expected))
cat(sprintf("relative difference = %.3e\n", abs(auc_inf / auc_expected - 1)))

stopifnot(abs(auc_inf / auc_expected - 1) < 1e-3)
```

## Check 3 - exact dose proportionality

The model is linear, so Cmax and AUC must scale exactly with dose across the
80 / 160 / 240 mg range the paper simulates in Figure 1.

```{r check-dose-prop}
dose_prop <- vapply(c(80, 160, 240), function(d) {
  ev <- rxode2::et(amt = d, cmt = "depot") |>
    rxode2::et(grid_fine, cmt = "central") |>
    as.data.frame()
  ev$CRCL <- 104.38
  ev$ALB  <- 49.0
  s <- rxode2::rxSolve(mod_tv, ev, returnType = "data.frame") |>
    dplyr::filter(!is.na(Cc))
  auc <- sum(diff(s$time) * (head(s$Cc, -1) + tail(s$Cc, -1)) / 2)
  c(cmax = max(s$Cc), auc24 = auc)
}, numeric(2))

prop_tab <- tibble(
  `Dose (mg)`            = c(80, 160, 240),
  `Cmax (ug/mL)`         = dose_prop["cmax", ],
  `AUC0-24 (ug*h/mL)`    = dose_prop["auc24", ],
  `Cmax / dose`          = dose_prop["cmax", ] / c(80, 160, 240),
  `AUC / dose`           = dose_prop["auc24", ] / c(80, 160, 240)
)
knitr::kable(prop_tab, digits = 6, caption = "Dose proportionality of the packaged model.")

stopifnot(
  max(abs(prop_tab$`Cmax / dose` / prop_tab$`Cmax / dose`[1] - 1)) < 1e-10,
  max(abs(prop_tab$`AUC / dose`  / prop_tab$`AUC / dose`[1]  - 1)) < 1e-10
)
```

## Check 4 - the covariate sub-model is the printed power model

`cl` returned by the solver must equal
`43.70 * (CRCL/104.38)^0.48 * (ALB_gdL/4.90)^-1.83` exactly, including at the
values the paper extrapolates to in its own dosing simulations.

```{r check-covariate}
cov_grid <- tidyr::crossing(CRCL = c(80, 104.38, 130), ALB_gdL = c(3.5, 4.90, 5.5))

ev_cov <- cov_grid |>
  dplyr::mutate(id = dplyr::row_number(), ALB = ALB_gdL * 10) |>
  tidyr::crossing(time = c(0, 1)) |>
  dplyr::mutate(
    amt  = ifelse(time == 0, 80, NA_real_),
    evid = ifelse(time == 0, 1L, 0L),
    cmt  = ifelse(time == 0, "depot", "central")
  ) |>
  dplyr::arrange(id, time)

sim_cov <- rxode2::rxSolve(mod_tv, ev_cov, keep = c("CRCL", "ALB"),
                           returnType = "data.frame")

cov_chk <- sim_cov |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::distinct(id, CRCL, ALB, cl) |>
  dplyr::mutate(
    ALB_gdL  = ALB * 0.1,
    expected = 43.70 * (CRCL / 104.38)^0.48 * (ALB_gdL / 4.90)^(-1.83),
    rel_diff = abs(cl / expected - 1)
  )

cov_chk |>
  dplyr::select(`CrCL (mL/min)` = CRCL, `Albumin (g/dL)` = ALB_gdL,
                `CL/F solved (L/h)` = cl, `CL/F expected (L/h)` = expected) |>
  knitr::kable(digits = 4, caption = "Solved CL/F against Equation (2) evaluated directly.")

stopifnot(max(cov_chk$rel_diff) < 1e-10)
```

## Check 5 - terminal half-life matches the disposition eigenvalue

```{r check-halflife}
# `Cc > 1e-12` guards log() against solver round-off in the far tail. This is a
# typical-value solve, not the PKNCA input.
tail_win <- sim_tv |> dplyr::filter(time >= 8, time <= 24, Cc > 1e-12)
lz <- -coef(lm(log(tail_win$Cc) ~ tail_win$time))[[2]]

cat(sprintf("terminal t1/2 from the beta eigenvalue = %.4f h\n", t_half_beta))
cat(sprintf("terminal t1/2 by log-linear regression = %.4f h\n", log(2) / lz))

stopifnot(abs((log(2) / lz) / t_half_beta - 1) < 0.01)
```

# Virtual cohort and stochastic simulation

Original subject-level data are not public. The cohort below draws
covariates from the Table S5 marginal distributions and simulates the three
single-dose levels the paper used for external validation (Table S6).

```{r cohort}
# set.seed() seeds R's RNG, not rxode2's; rxode2's streams are partitioned per
# solver thread, so this cohort differs between a 2-core CI runner and a
# 16-thread workstation. Every assertion below is written to hold for any
# cohort the model can produce (pattern 12 of known-vignette-failure-patterns).
set.seed(20230161)

n_per_arm <- 100L   # well under the 200-per-arm cap

make_arm <- function(dose, id_offset) {
  subj <- tibble(
    id      = id_offset + seq_len(n_per_arm),
    dose_mg = dose,
    # Table S5: CrCL 107.53 +/- 17.28 mL/min, albumin 4.92 +/- 0.18 g/dL.
    CRCL    = pmin(pmax(rnorm(n_per_arm, 107.53, 17.28), 60), 165),
    ALB     = pmin(pmax(rnorm(n_per_arm, 49.2, 1.8), 42), 57)  # SI g/L
  )
  obs_times <- sort(unique(c(seq(0, 2, by = 0.05), seq(2.25, 12, by = 0.25),
                             seq(13, 24, by = 1))))
  dplyr::bind_rows(
    subj |> dplyr::mutate(time = 0, amt = dose, evid = 1L, cmt = "depot"),
    subj |> tidyr::crossing(time = obs_times) |>
      dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
  ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  make_arm(80,  0L),
  make_arm(160, 100L),
  make_arm(240, 200L)
) |>
  dplyr::mutate(treatment = paste0(dose_mg, " mg"))

# Each (id, time, evid) triple must be unique. Note the pre-dose record is a
# genuine pair -- id X has a dose (evid 1) and an observation (evid 0) both at
# t = 0 -- so evid is part of the key. Do NOT wrap this in unique(), which
# would make the assertion vacuously true.
stopifnot(
  !anyDuplicated(events[, c("id", "time", "evid")]),
  nrow(events) == 3L * n_per_arm * (1L + 93L),   # 1 dose + 93 observation times
  identical(sort(unique(events$treatment)), c("160 mg", "240 mg", "80 mg"))
)
```

```{r simulate}
sim <- rxode2::rxSolve(mod, events = events,
                       keep = c("treatment", "dose_mg", "CRCL", "ALB")) |>
  as.data.frame()
```

```{r figure-1-shape, fig.width = 7, fig.height = 4}
# Shape of the packaged model across the three single-dose levels the paper
# simulates in its Figure 1. NOTE: this reproduces the SHAPE OF THE PACKAGED
# PARAMETER SET, not the observed data overlaid in the paper's figure -- see
# the Errata section.
# The `time >= 0.05` bound drops only the pre-dose record, which is 0 and so
# cannot be drawn on the log scale below. It is NOT a PKNCA input filter --
# the PKNCA chunk filters on !is.na(Cc) alone and adds its own time-zero row.
sim |>
  dplyr::filter(!is.na(Cc), time >= 0.05) |>
  dplyr::group_by(treatment, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~treatment) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Zaltoprofen concentration (ug/mL)",
       title = "Packaged model, single oral doses",
       caption = "Median and 5th-95th percentile of 100 simulated subjects per arm.")
```

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time = 0 row per (id, treatment); pre-dose Cc = 0 is correct for
# an extravascular single dose.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

stopifnot(nrow(sim_nca) > 0, all(sim_nca$Cc >= 0))

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_summary <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median)

nca_summary |>
  dplyr::select(treatment, cmax, tmax, aucinf.obs, half.life) |>
  dplyr::rename(
    "Dose"                    = treatment,
    "Cmax (ug/mL)"            = cmax,
    "Tmax (h)"                = tmax,
    "AUC0-inf (ug*h/mL)"      = aucinf.obs,
    "t1/2 (h)"                = half.life
  ) |>
  knitr::kable(digits = 3,
               caption = "Median NCA parameters of the packaged model, by dose.")
```

```{r pknca-gate}
# The cohort median AUC must sit near Dose/(CL/F) evaluated at the cohort's
# median covariates. This is a distributional statement, so the bound is on the
# CENTRE and is loose enough to admit any cohort the model can draw, while
# still going red on a mis-transcribed clearance (which moves it 6-7 fold).
med <- nca_summary |> dplyr::mutate(dose_mg = as.numeric(sub(" mg", "", treatment)))
cl_ref <- 43.70 * (median(events$CRCL) / 104.38)^0.48 *
  (median(events$ALB) * 0.1 / 4.90)^(-1.83)
med$expected_auc <- med$dose_mg / cl_ref
med$pct_diff <- 100 * (med$aucinf.obs / med$expected_auc - 1)

knitr::kable(
  med |> dplyr::select(`Dose` = treatment,
                       `Median AUC0-inf` = aucinf.obs,
                       `Dose / (CL/F) at cohort median covariates` = expected_auc,
                       `% difference` = pct_diff),
  digits = 3,
  caption = "Cohort median AUC against the Dose/CL identity."
)

stopifnot(
  max(abs(med$pct_diff)) < 15,        # centre; log-normal CL skew accounts for a few %
  all(abs(med$half.life / t_half_beta - 1) < 0.10)
)
```

## Comparison against published NCA

**The source paper publishes no NCA parameter table for the modelling
cohort**, so `nlmixr2lib::ncaComparisonTable()` has no reference to compare
against. The paper has no numbered main-text tables at all; Figure S1 plots
NCA parameters by CYP2C9 genotype but reports only p-values, and Table S6
lists the external studies' designs without their results. The nearest
published numbers are the steady-state concentration ratios in Results 2.4,
used below.

## The covariate sub-model against the paper's own published ratios

Results 2.4 reports simulated steady-state mean concentrations for six
covariate scenarios relative to the "normal" group (CrCL 104.38 mL/min,
albumin 4.90 g/dL), and it prints each one as a **ratio** to that group. The
structural scale therefore cancels out entirely: these test the covariate
exponents and the centring constants alone. This is the one place the
packaged model can be checked against the paper's own numbers, and it is the
evidence that the defect described in the Errata is confined to the
structural parameters.

All six ratios below are quoted directly from Results 2.4 rather than
divided out of the concentration pairs it also prints (e.g. the composite
CrCL 130 + albumin 3.5 scenario is stated as "approximately 0.43 times lower
than that of the normal group", so 0.43 is used rather than the rounded
0.19 / 0.43 = 0.44 the printed concentrations would give).

```{r covariate-vs-published}
cav_ratio <- function(crcl, alb_gdL) {
  1 / ((crcl / 104.38)^0.48 * (alb_gdL / 4.90)^(-1.83))
}

pub <- tibble::tribble(
  ~Scenario,                          ~crcl,   ~alb, ~published, ~source,
  "CrCL 130 mL/min, albumin median",   130,    4.90, 0.87,       "Results 2.4: 'CrCL 130 and 80 mL/min groups were 0.87 and 1.30 times'",
  "CrCL 80 mL/min, albumin median",     80,    4.90, 1.30,       "Results 2.4: same sentence",
  "CrCL median, albumin 5.5 g/dL",     104.38, 5.50, 1.28,       "Results 2.4: 'albumin levels of 5.5 and 3.5 g/dL were 1.28 and 0.48 times'",
  "CrCL median, albumin 3.5 g/dL",     104.38, 3.50, 0.48,       "Results 2.4: same sentence",
  "CrCL 130 mL/min, albumin 3.5 g/dL", 130,    3.50, 0.43,       "Results 2.4 and Discussion: 'approximately 0.43 times lower than that of the normal group'",
  "CrCL 80 mL/min, albumin 5.5 g/dL",   80,    5.50, 1.49,       "Results 2.4 and Discussion: 'increased by approximately 1.49 times compared to the normal group'"
) |>
  dplyr::mutate(
    model    = cav_ratio(crcl, alb),
    pct_diff = 100 * (model / published - 1)
  )

pub |>
  dplyr::select(Scenario,
                `Published ratio` = published,
                `Model ratio` = model,
                `% difference` = pct_diff,
                `Source` = source) |>
  knitr::kable(digits = 3,
               caption = "Steady-state exposure ratios relative to the normal group. Ratios are independent of the structural scale.")

# Deterministic (no cohort draw), so a tight bound is appropriate. Realised
# max |diff| is 13.07% (the CrCL 130 + albumin 3.5 composite).
#
# What this gate does and does not discriminate, measured on this same six-row
# anchor set:
#   as printed .................  13.1%  (passes)
#   CrCL exponent sign flipped ..  39.6%  (fails)
#   albumin exponent sign flipped 287.4%  (fails)
#   both exponents swapped ...... 308.4%  (fails)
#   Table S5 MEANS (107.53 / 4.92) as the centring constants
#                                 13.8%  (PASSES -- see below)
# So the gate is decisive about the exponents and their signs, but it does NOT
# separate the Results 2.4 medians from the Table S5 means: the two centrings
# differ by only 3% and 0.4% respectively, which is far inside the two-
# significant-figure rounding of the published ratios. The medians are used
# because Eq. (2) writes mCrCL / mAlbumin and the text defines m as the median
# -- that is a source-trace argument, not something this check proves.
stopifnot(
  max(abs(pub$pct_diff)) < 15,
  nrow(pub) == 6L                 # all six published ratios are exercised
)
```

All six published ratios are reproduced to within 13.1%, with the residual
attributable to the paper reporting its concentrations to two significant
figures and to its ratios being means over a simulated cohort rather than
typical values. **The covariate half of the model is therefore corroborated
by the paper's own numbers; the structural half is not.**

# Errata: the Table S3 versus figures conflict {#errata-the-table-s3-versus-figures-conflict}

## The demonstration

The argument needs no digitisation and no fitting. `AUC(0-inf) = Dose / CL`
is parameterisation-free for any linear model with first-order input, so the
printed `tv CL/F = 43.70 L/h` fixes the 80 mg exposure at
`80 / 43.70 = 1.83 ug*h/mL` regardless of every other parameter. Solving the
full printed set gives:

```{r errata-summary}
i <- which.max(sim_tv$Cc)
errata_tab <- tibble(
  Quantity = c("Cmax (ug/mL)", "Tmax (h)", "AUC0-24 (ug*h/mL)", "terminal t1/2 (h)",
               "steady-state mean, 80 mg q24h (ug/mL)"),
  `Packaged (printed Table S3)` = c(sim_tv$Cc[i], sim_tv$time[i], auc_trap,
                                    t_half_beta, 80 / (43.70 * 24)),
  `Paper's own figures / text` = c("~3.6 (Fig. 1A observed mean)",
                                   "~1.0-1.5 (Fig. 1A)",
                                   "~12.5 (Fig. 1A, trapezoid)",
                                   "~6 (Fig. 1A terminal slope)",
                                   "0.43 (Results 2.4, printed)")
)
knitr::kable(errata_tab, digits = 4,
             caption = "Packaged parameter set against the same paper's own figures and text.")
```

The last row is the strongest single item because it is a **printed number,
not a digitised one**: Results 2.4 states the normal group's steady-state
mean plasma concentration is 0.43 ug/mL for 80 mg, while
`Dose / (CL/F * tau) = 80 / (43.70 * 24) = 0.076 ug/mL` -- a 5.6-fold gap.
(At the 8 h interval the model gives 0.229 ug/mL, still 1.9-fold low, and the
paper states its 8 h and 24 h comparative results were equivalent.)

## No reading of Table S3 reconciles it

The alternatives below are evaluated live from the closed-form solution, so
the rejection is reproducible rather than asserted. The target column is the
paper's own Figure 1A behaviour (Cmax about 3.6 ug/mL, Tmax about 1 h,
AUC(0-24) about 12.5 ug*h/mL, terminal t1/2 about 6 h).

```{r errata-hypotheses}
summarise_set <- function(label, ka, vc, cl, vp, q) {
  tt <- sort(unique(c(seq(0, 4, by = 0.002), seq(4, 24, by = 0.01))))
  cc <- cf_2cmt_oral(tt, dose = 80, ka = ka, vc = vc, cl = cl, vp = vp, q = q)
  k10 <- cl / vc; k12 <- q / vc; k21 <- q / vp
  b <- k10 + k12 + k21
  beta <- (b - sqrt(b^2 - 4 * k10 * k21)) / 2
  i <- which.max(cc)
  tibble(
    Hypothesis         = label,
    `Cmax (ug/mL)`     = cc[i],
    `Tmax (h)`         = tt[i],
    `AUC0-24 (ug*h/mL)`= sum(diff(tt) * (head(cc, -1) + tail(cc, -1)) / 2),
    `t1/2 beta (h)`    = log(2) / beta
  )
}

hyp <- dplyr::bind_rows(
  summarise_set("printed as-is",                        1.73, 4.88,  43.70, 40.56, 5.61),
  summarise_set("CL/F decimal shift (4.37)",            1.73, 4.88,   4.37, 40.56, 5.61),
  summarise_set("CL/F and CL2/F values swapped",        1.73, 4.88,   5.61, 40.56, 43.70),
  summarise_set("V/F and V2/F values swapped",          1.73, 40.56, 43.70,  4.88, 5.61),
  summarise_set("both pairs swapped",                   1.73, 40.56,  5.61,  4.88, 43.70),
  summarise_set("CL/F read as mL/min (2.622 L/h)",      1.73, 4.88,   2.622, 40.56, 5.61)
)

knitr::kable(hyp, digits = 3,
             caption = "Alternative readings of Table S3. Target: Cmax ~3.6, Tmax ~1, AUC0-24 ~12.5, t1/2 ~6.")
```

Every alternative fails on at least one of the four targets, and the two that
come closest on AUC (the decimal shift and the clearance swap) miss Tmax by
2-4 fold because `V/F = 4.88 L` is itself too small: at the printed values
`k10 = CL/V = 8.95 /h` is five times larger than `ka = 1.73 /h`, which forces
an early spike and collapse no matter what the clearance is. **At least two
of the five structural values would have to change**, so this is not a single
repairable typo and it was not repaired here.

## What passes, and why the table still looks trustworthy

Every internal-consistency check on Table S3 passes, which is why the defect
survives casual inspection:

* `AIC = -2LL + 2 * nParameter` holds on every row of Table S1
  (e.g. `168.75 + 2*11 = 190.75`).
* The parameter count inverts exactly: 9 for the base model (5 structural,
  3 IIV, 1 residual) and 11 for the final model (+2 covariate effects),
  matching Table S2's `nParameter` column and the 11 rows of Table S3.
* The `IIV (%)` column equals `sqrt(omega^2) * 100` on the unrounded variances
  for all three etas (checked live above).
* Table S4's 1000-replicate bootstrap medians sit within 20% of every point
  estimate, and every RSE is plausible.
* The covariate sub-model reproduces the paper's own published ratios to
  within 13% (checked live above).

So Table S3 is a coherent, self-consistent set of numbers from a real Phoenix
run. What it is not is consistent with the exposure the same paper plots and
quotes. The most likely explanations -- none of which can be chosen from the
published material -- are that Table S3 reports a different run than the one
used for Figures 1-5, or that a units convention differs between the fit and
the table.

## Figure 1A digitised values (supporting context only)

For completeness, the values below were digitised from a 400 dpi render of
Figure 1A (80 mg single dose) during the extraction. **They are
non-paper-derived, approximate, and no assertion in this vignette depends on
them** -- the demonstration above rests on the printed 0.43 ug/mL and on the
`AUC = Dose/CL` identity.

| Time (h) | Fig. 1A observed mean (ug/mL) | Packaged model (ug/mL) | Observed / predicted |
|---|---|---|---|
| 0.25 | 2.0 | 1.94 | 1.0x |
| 0.5 | 2.9 | 1.41 | 2.1x |
| 1 | 3.6 | 0.62 | 5.8x |
| 2 | 3.0 | 0.127 | 24x |
| 4 | 0.95 | 0.021 | 46x |
| 8 | 0.15 | 0.011 | 14x |
| 12 | 0.11 | 0.0065 | 17x |
| 24 | 0.08 | 0.0015 | 54x |

# Assumptions and deviations

* **The packaged parameter set does not reproduce the source paper's own
  figures or its quoted steady-state concentrations.** Values are shipped
  verbatim from supplementary Table S3 per the standing rule that printed
  values have authority over figures (operator ruling, 2026-09-02, in
  preference to deferring for author correspondence). The full quantitative
  demonstration is in the Errata section above. This vignette gates on
  self-consistency and on the paper's published covariate ratios; it
  deliberately does **not** gate on reproducing Figures 1-2.
* **Between-subject variances** are back-calculated from Table S3's
  four-significant-figure `IIV (%)` column rather than read from its
  two-decimal `omega^2` column, because the two encode the same quantity and
  the former is more precise. The `omega = IIV(%)/100` relation is verified
  live in the `iiv-precision-check` chunk, and the competing log-normal
  reading is shown to be inconsistent with the printed variances.
* **Centring constants** use the medians stated in Results 2.4 and the
  Figure 2-5 captions (CrCL 104.38 mL/min, albumin 4.90 g/dL), not the
  Table S5 means (107.53 mL/min, 4.92 g/dL). This rests on the source trace
  alone: Equation (2) writes `mCrCL` and `mAlbumin`, and Results 2.4 states
  the normal group "had CrCL and albumin levels of 104.38 mL/min and
  4.90 g/dL ... meaning the group with median values". The published-ratio
  check does **not** corroborate the choice -- the two centrings differ by
  only 3% and 0.4%, so both readings pass that gate (13.1% versus 13.8%
  maximum deviation). No check available in this vignette can distinguish
  them; the medians are used because the paper says median.
* **Albumin units.** The canonical `ALB` column is SI g/L, while Jang 2023
  calibrated its exponent against US-convention g/dL, so `model()` applies
  `alb_gdL <- ALB * 0.1` before the ratio. Supply albumin in g/L
  (4.90 g/dL = 49.0 g/L).
* **Creatinine clearance** is raw Cockcroft-Gault mL/min, **not** normalised
  to 1.73 m^2. The cohort was entirely male, so no 0.85 female multiplier
  was applied when the source computed it.
* **CYP2C9 genotype was screened and rejected**, so it is recorded in
  `covariatesDataExcluded` rather than `covariateData`; no coefficient is
  reported for it anywhere in the paper, so it could not be encoded even
  optionally. BSA, BMI, ALT, ALP and BUN were also screened and rejected and
  are recorded the same way.
* **No bioavailability parameter.** All volumes and clearances are apparent
  (`X/F`); the paper reports no absolute bioavailability and none is
  estimable from single-route oral data.
* **The virtual cohort is synthetic.** Subject-level covariates are drawn
  from the Table S5 marginal normal distributions (truncated to plausible
  physiological ranges); the paper publishes no joint distribution and no
  individual data. The 160 mg and 240 mg arms use the same covariate
  distribution as the 80 mg arm, whereas the paper's Figure 1 external
  validation drew those doses from different published studies (Table S6).
* **The Figure 1A digitised values** in the Errata section are
  non-paper-derived (operator digitisation from a 400 dpi render) and are
  presented as supporting context only; no assertion depends on them.
