---
title: "Itraconazole nanocrystal formulation (Jansen 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Itraconazole nanocrystal formulation (Jansen 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Jansen_2023_itraconazole"))$reference`
* Description: `r rxode2::rxode(readModelDb("Jansen_2023_itraconazole"))$description`
* Article: <https://doi.org/10.1093/jac/dkad072>
* Supplement (figures S1-S5 and the final model NONMEM control stream):
  <https://www.ebi.ac.uk/europepmc/webservices/rest/PMC10154123/supplementaryFiles>

# Population

Ten adults (median age 47.5 years, range 22.0-59.0; 50% female; median weight
78.3 kg, range 60.0-92.5 kg) received a matched allogeneic bone marrow
transplant after conditioning with idarubicin, cyclophosphamide and total body
irradiation, and took part in a prospective open-label Phase II study of
itraconazole nanocrystal formulation (NCF) for prophylaxis of invasive fungal
disease at Radboud University Medical Center, Nijmegen (Jansen 2023, Table 1).
Underlying diseases were acute lymphatic leukaemia (30%), acute myeloid
leukaemia (20%), chronic myelomonocytic leukaemia (20%), non-Hodgkin lymphoma
(20%) and myelofibrosis (10%).

Every subject received itraconazole NCF 200 mg as a 2 h intravenous infusion
twice daily for 2 days, then 200 mg once daily until day 14. Full
pharmacokinetic curves were drawn on days 7 and 14, with pre- and post-infusion
samples until day 6, pre-infusion samples on days 10 and 12, and washout
samples on days 16, 17, 18, 19 and 28. This yielded 471 itraconazole and 471
paired hydroxy-itraconazole concentrations, all above the LLOQ. The study was
explicitly not powered to identify covariates, so body weight - entered a
priori as fixed allometric scaling - is the only covariate in the model.

The same information is available programmatically via
`readModelDb("Jansen_2023_itraconazole")()$population`.

# Model structure

The published model is a semi-mechanistic parent-metabolite model
(Jansen 2023 Figure S2, and the `$MODEL` / `$PK` blocks of the supplementary
final model control stream):

* The infusion enters a **nanocrystal-bound itraconazole** compartment
  (`central_np`), from which drug dissolves by a first-order rate constant
  `kN` fixed at 4.62 /h (a 9 min dissolution half-life in human plasma).
* Dissolved itraconazole follows **two-compartment** disposition
  (`central`, `peripheral1`).
* The fraction of itraconazole metabolised to hydroxy-itraconazole was
  **fixed to 1**, so the whole of `CL_P` feeds the one-compartment
  hydroxy-itraconazole model (`central_ohi`); there is no `K10` in the control
  stream.
* The nanocrystal-bound volume `V_N` is set equal to the itraconazole central
  volume `V_P1` (an explicit assumption of the paper, made in the absence of
  nanocrystal-bound concentration measurements).
* The **observed** itraconazole concentration is the sum of the
  nanocrystal-bound and dissolved concentrations
  (`$ERROR: IPRED = LOG(C1 + C4)`), which is what `Cc` returns here.

All concentrations in the source analysis were converted to molar equivalents
before fitting, so that the fixed `fm = 1` conversion is mole-for-mole. The
packaged model therefore works in **mmol** and **mmol/L**, matching the y-axis
of Jansen 2023 Figure 1. This vignette converts to mg/L for comparison against
the paper's reported trough concentrations and therapeutic-drug-monitoring
targets.

```{r constants}
# Molecular weights are chemical constants, NOT taken from the paper (the paper
# does not print them). Itraconazole C35H38Cl2N8O4 = 705.6 g/mol (PubChem CID
# 55283); hydroxy-itraconazole C35H38Cl2N8O5 = 721.6 g/mol (PubChem CID 108222).
MW_ITZ <- 705.6
MW_OHI <- 721.6

# 200 mg itraconazole per infusion, expressed in the model's molar dose unit.
dose_mmol <- 200 / MW_ITZ
dose_mmol
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Jansen_2023_itraconazole.R`. The table
below collects them in one place for review. "Control stream" refers to the
`FINAL MODEL CONTROL STREAM` printed at the end of the supplementary data
document.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL_P)         | 4.29 L/h | Table 2 (95% CI 3.85-4.80); control stream `$THETA` TVCLP |
| `lvc` (V_P1 = V_N)   | 14.1 L | Table 2 (95% CI 11.8-17.2); control stream `$THETA` TVVP1/TVVN |
| `lq` (Q_P)           | 53.0 L/h | Table 2 (95% CI 45.3-63.4); control stream `$THETA` TVQP |
| `lvp` (V_P2)         | 1660 L | Table 2 (95% CI 1558-1775); control stream `$THETA` TVVP2 |
| `lcl_ohi` (CL_M)     | 2.86 L/h | Table 2 (95% CI 2.43-3.33); control stream `$THETA` TVCLM |
| `lvc_ohi` (V_M)      | 43.1 L | Table 2 (95% CI 36.9-50.4); control stream `$THETA` TVVM |
| `lkdiss` (kN)        | 4.62 /h, fixed | Methods, "Population pharmacokinetic analysis"; control stream `$PK KN = 4.62` |
| `e_wt_cl`            | 0.75, fixed | Methods, "a fixed exponent of 3/4 for (intercompartmental) clearance"; control stream `(WT/70)**0.75` |
| `e_wt_vc`            | 1.0, fixed | Methods, "and 1 for volumes of distribution"; control stream `(WT/70)**1` |
| `etalcl`             | var 0.0126 | Control stream `$OMEGA` "IIV CLP"; Table 2 reports 11.3% = sqrt(exp(0.0126) - 1) |
| `etalcl_ohi`         | var 0.0506 | Control stream `$OMEGA` "IIV CLM"; Table 2 reports 22.8% = sqrt(exp(0.0506) - 1) |
| `expSd`              | sqrt(0.214) | Control stream `$SIGMA BLOCK(2)` first diagonal; Table 2 Error_P 48.8% = sqrt(exp(0.214) - 1) |
| `expSd_ohi`          | sqrt(0.0358) | Control stream `$SIGMA BLOCK(2)` second diagonal; Table 2 Error_M 19.1% = sqrt(exp(0.0358) - 1) |
| `d/dt(central_np)`   | n/a | Control stream `$PK K41 = KN`; dose administered into `COMP=(NANO)` |
| `d/dt(central)`, `d/dt(peripheral1)` | n/a | Control stream `$PK K12 = QP/VP1`, `K21 = QP/VP2`, `K13 = CLP/VP1` (no K10) |
| `d/dt(central_ohi)`  | n/a | Control stream `$PK K13 = CLP/VP1`, `K30 = CLM/VM`; Methods, fm assumed 1 |
| `Cc = (central + central_np)/vc` | n/a | Control stream `$ERROR: C1 = A(1)/VP1`, `C4 = A(4)/VN`, `IPRED = LOG(C1+C4)` |
| `Cc_ohi = central_ohi/vc_ohi` | n/a | Control stream `$ERROR: C3 = A(3)/VM`, `IPRED = LOG(C3)`; `$PK S3 = VM` |

# Dosing regimens used below

```{r regimens}
mod <- readModelDb("Jansen_2023_itraconazole")

# Trial protocol: 200 mg NCF over 2 h, q12h x 4 doses (days 1-2), then q24h
# until day 14 (12 further doses, last at t = 312 h).
trial_doses <- rxode2::et(amt = dose_mmol, dur = 2, ii = 12, addl = 3,
                          cmt = "central_np") |>
  rxode2::et(amt = dose_mmol, dur = 2, time = 48, ii = 24, addl = 11,
             cmt = "central_np")

# Observation rows name an ODE state in `cmt` and select the model endpoint
# with `dvid`. rxode2 requires an observation record of a multi-endpoint model
# to resolve to one of the endpoint compartments; `dvid` does that without
# writing an algebraic observable name into `cmt` (which would inject a
# compartment slot and renumber the states). Every algebraic observable
# (`Cc`, `Cc_np`, `Cc_ohi`) is returned as a column regardless of `dvid`.
obs_rows <- function(times) {
  data.frame(time = times, evid = 0L, amt = NA_real_,
             cmt = "central", dvid = 1L)
}

add_obs <- function(doses, times) {
  bind_rows(as.data.frame(doses), obs_rows(times)) |>
    arrange(time)
}
```

# Replicate Figure 1

Figure 1 of Jansen 2023 shows simulated nanocrystal-bound itraconazole (dotted),
itraconazole (solid) and hydroxy-itraconazole (dashed) for a typical individual
given the study protocol, on a log y-axis in mmol/L over 144 h. The
nanocrystal-bound curve spikes and collapses within minutes of each infusion
(dissolution half-life 9 min), so it appears as a near-vertical excursion at
each dose.

```{r figure-1, fig.cap = "Replicates Figure 1 of Jansen 2023: typical-individual (70 kg) profiles over the first 6 days of the study protocol."}
ev_typ <- add_obs(trial_doses, seq(0, 144, by = 0.05)) |>
  mutate(WT = 70)

sim_typ <- rxode2::rxSolve(mod, ev_typ, omega = NA, addDosing = FALSE,
                           returnType = "data.frame")

fig1 <- sim_typ |>
  transmute(
    time,
    `Nanocrystal-bound itraconazole` = Cc_np,
    `Itraconazole (dissolved)`       = Cc - Cc_np,
    `Hydroxy-itraconazole`           = Cc_ohi
  ) |>
  pivot_longer(-time, names_to = "Analyte", values_to = "conc") |>
  filter(conc > 0)

ggplot(fig1, aes(time, conc, linetype = Analyte)) +
  geom_line() +
  scale_y_log10(limits = c(1e-4, 1e-2)) +
  scale_x_continuous(breaks = seq(0, 144, by = 24)) +
  scale_linetype_manual(values = c(
    "Nanocrystal-bound itraconazole" = "dotted",
    "Itraconazole (dissolved)"       = "solid",
    "Hydroxy-itraconazole"           = "dashed"
  )) +
  labs(x = "Time (h)", y = "Concentration (mmol/L)", linetype = NULL,
       caption = "Replicates Figure 1 of Jansen 2023.") +
  theme(legend.position = "bottom")
```

The reproduced curves match the published panel in both level and shape: the
itraconazole trough climbs from roughly 1.5e-4 mmol/L after the first
inter-dose interval to roughly 9e-4 mmol/L by 144 h, while hydroxy-itraconazole
rises smoothly past the parent from about 40 h onward.

```{r figure-1-checks}
tr <- function(t) sim_typ[which.min(abs(sim_typ$time - t)), ]
# Deterministic (typical-value) quantities: a tight bound is appropriate.
stopifnot(
  # Itraconazole trough at 144 h, read off Figure 1 as ~9e-4 mmol/L.
  abs(tr(144)$Cc / 9.0e-4 - 1) < 0.15,
  # Hydroxy-itraconazole overtakes itraconazole partway through the profile
  # and stays above it thereafter, as in Figure 1.
  tr(24)$Cc_ohi  > tr(24)$Cc,
  tr(144)$Cc_ohi > tr(144)$Cc,
  # The nanocrystal-bound term is spent long before the next dose: at the 144 h
  # trough it contributes nothing measurable to the observed itraconazole.
  tr(144)$Cc_np / tr(144)$Cc < 1e-6
)
```

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the trial's body-weight distribution (median 78.3 kg, observed range
60.0-92.5 kg; Jansen 2023 Table 1) and uses the model's own IIV on itraconazole
and hydroxy-itraconazole clearance.

```{r cohort}
# `set.seed()` seeds R's RNG, not rxode2's, and rxode2's streams are
# partitioned per solver thread -- so this cohort differs between a 2-core CI
# runner and a 16-thread workstation. Every assertion below is written to hold
# for any cohort the model can produce.
set.seed(20230501)

n_sub <- 100  # well under the 200-per-arm cap

cohort <- tibble(
  id = seq_len(n_sub),
  # Log-normal about the published median; SD chosen so the central 95% of the
  # distribution spans roughly the published 60.0-92.5 kg range.
  WT = 78.3 * exp(rnorm(n_sub, 0, 0.12)),
  treatment = "200 mg NCF IV"
)

base_ev <- add_obs(trial_doses, seq(0, 336, by = 0.5))

events <- tidyr::expand_grid(cohort, base_ev) |>
  arrange(id, time)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

```{r simulate}
sim <- rxode2::rxSolve(mod, events, keep = c("WT", "treatment"),
                       addDosing = FALSE, returnType = "data.frame")
if (is.null(sim$id)) sim$id <- 1L
```

# Steady-state trough concentrations

Jansen 2023 reports a mean (SD) trough concentration at steady state (after day
7) of 0.79 mg/L (0.35) for itraconazole and 1.31 mg/L (0.29) for
hydroxy-itraconazole. Pre-dose samples in that window were drawn on days 7, 10,
12 and 14, i.e. at 144, 216, 264 and 312 h after the first infusion.

The primary comparison uses the **typical-value** profile at the study's median
weight, which is a deterministic quantity and can therefore be gated tightly.

```{r trough-typical}
trough_times <- c(144, 216, 264, 312)

ev_med <- add_obs(trial_doses, trough_times) |>
  mutate(WT = 78.3)

sim_med <- rxode2::rxSolve(mod, ev_med, omega = NA, addDosing = FALSE,
                           returnType = "data.frame")

trough_cmp <- tibble(
  Analyte    = c("Itraconazole", "Hydroxy-itraconazole"),
  Published  = c(0.79, 1.31),
  Simulated  = c(mean(sim_med$Cc) * MW_ITZ, mean(sim_med$Cc_ohi) * MW_OHI)
) |>
  mutate(`% diff` = 100 * (Simulated - Published) / Published)

trough_cmp |>
  rename("Published mean Cmin (mg/L)" = Published,
         "Simulated typical Cmin (mg/L)" = Simulated) |>
  knitr::kable(digits = c(0, 2, 3, 1),
               caption = "Mean trough concentration over days 7, 10, 12 and 14 at the study median weight (78.3 kg), against the observed means of Jansen 2023 Results.")
```

```{r trough-checks}
# Structural gate: a mis-transcribed clearance, dose, molecular weight or unit
# moves these by tens of percent. The typical-value solve is deterministic, so
# the bound does not have to absorb cohort noise.
stopifnot(max(abs(trough_cmp$`% diff`)) < 10)
```

For context the same window is summarised over the virtual cohort. Note that
these are individual predictions: the published means are arithmetic means of
*observations*, which the log-scale residual error would inflate by
`exp(sigma^2 / 2)` (about 11% for itraconazole and 2% for
hydroxy-itraconazole) relative to a noise-free prediction.

```{r trough-cohort}
cohort_troughs <- sim |>
  filter(time %in% trough_times) |>
  summarise(
    .by = id,
    ITZ = mean(Cc) * MW_ITZ,
    OHI = mean(Cc_ohi) * MW_OHI
  )

tibble(
  Analyte = c("Itraconazole", "Hydroxy-itraconazole"),
  Median  = c(median(cohort_troughs$ITZ), median(cohort_troughs$OHI)),
  P05     = c(quantile(cohort_troughs$ITZ, 0.05), quantile(cohort_troughs$OHI, 0.05)),
  P95     = c(quantile(cohort_troughs$ITZ, 0.95), quantile(cohort_troughs$OHI, 0.95))
) |>
  rename("Median (mg/L)" = Median, "5th pct" = P05, "95th pct" = P95) |>
  knitr::kable(digits = 3,
               caption = "Virtual-cohort mean trough over days 7, 10, 12 and 14 (individual predictions).")
```

```{r trough-cohort-checks}
# Cohort-derived: assert the centre and a robust quantile, never an extreme.
stopifnot(
  abs(median(cohort_troughs$ITZ) / 0.79 - 1) < 0.25,
  abs(median(cohort_troughs$OHI) / 1.31 - 1) < 0.25
)
```

# PKNCA validation

## Steady-state exposure over the day-14 dosing interval

`PKNCA` is used for every NCA quantity. The paper reports no NCA table, so the
values below characterise the packaged model rather than reproducing a
published table.

```{r pknca-day14}
tau      <- 24
start_ss <- 312          # last dose of the protocol
end_ss   <- start_ss + tau

nca_conc <- function(df, col) {
  out <- df |>
    filter(!is.na(.data[[col]]), time >= start_ss, time <= end_ss) |>
    transmute(id, time, Cc = .data[[col]], treatment)
  out
}

intervals_ss <- data.frame(
  start = start_ss, end = end_ss,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, cav = TRUE, auclast = TRUE
)

dose_df <- events |>
  filter(evid == 1) |>
  transmute(id, time, amt, treatment)

nca_for <- function(col, concu) {
  conc_obj <- PKNCA::PKNCAconc(nca_conc(sim, col), Cc ~ time | treatment + id,
                               concu = concu, timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                               doseu = "mmol")
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals_ss))
}

nca_itz <- nca_for("Cc", "mmol/L")
nca_ohi <- nca_for("Cc_ohi", "mmol/L")
```

```{r pknca-day14-table}
nca_tbl <- function(res, analyte, mw) {
  as.data.frame(res$result) |>
    filter(PPTESTCD %in% c("cmax", "tmax", "cmin", "cav", "auclast")) |>
    mutate(
      value = if_else(PPTESTCD == "tmax", PPORRES, PPORRES * mw),
      Analyte = analyte
    ) |>
    summarise(.by = c(Analyte, PPTESTCD),
              Median = median(value),
              P05 = quantile(value, 0.05),
              P95 = quantile(value, 0.95)) |>
    mutate(Parameter = nlmixr2lib::ncaParamLabel(PPTESTCD),
           Units = if_else(PPTESTCD == "tmax", "h",
                           if_else(PPTESTCD == "auclast", "mg*h/L", "mg/L"))) |>
    select(Analyte, Parameter, Units, Median, P05, P95)
}

bind_rows(nca_tbl(nca_itz, "Itraconazole", MW_ITZ),
          nca_tbl(nca_ohi, "Hydroxy-itraconazole", MW_OHI)) |>
  rename("5th pct" = P05, "95th pct" = P95) |>
  knitr::kable(digits = 3,
               caption = "PKNCA over the day-14 dosing interval (312-336 h) of the virtual cohort, converted from molar to mass units.")
```

## Exact internal identities at true steady state

The terminal half-life implied by the published disposition parameters is
`log(2) / beta` with `beta` from the two-compartment micro-constants; it is
long enough (roughly 12 days) that the 14-day protocol has not reached true
steady state, which is why the troughs above are still climbing at day 14. At
true steady state the model satisfies three exact mass-balance identities,
which are checked here against `PKNCA`'s AUC over one dosing interval:

* `AUCtau(Cc - Cc_np) = Dose / CL_P` - every molecule leaving the dissolved
  itraconazole compartment does so through `CL_P`.
* `AUCtau(Cc_np) = Dose / (kN * V_P1)` - the nanocrystal-bound compartment has
  dissolution as its only exit.
* `AUCtau(Cc_ohi) = Dose / CL_M` - `fm` is fixed to 1.

```{r ss-identities}
n_days <- 120  # about 10 terminal half-lives
t_last <- (n_days - 1) * 24

ev_ss <- add_obs(
  rxode2::et(amt = dose_mmol, dur = 2, ii = 24, addl = n_days - 1,
             cmt = "central_np"),
  seq(t_last, t_last + tau, by = 0.05)
) |>
  mutate(WT = 70, id = 1L, treatment = "200 mg QD to steady state")

sim_ss <- rxode2::rxSolve(mod, ev_ss, omega = NA, addDosing = FALSE,
                          maxsteps = 200000L, atol = 1e-12, rtol = 1e-10,
                          returnType = "data.frame")
if (is.null(sim_ss$id)) sim_ss$id <- 1L
sim_ss$treatment <- "200 mg QD to steady state"

auc_ss <- function(col) {
  conc <- sim_ss |> transmute(id, time, Cc = .data[[col]], treatment)
  conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ time | treatment + id,
                               concu = "mmol/L", timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(
    tibble(id = 1L, time = t_last, amt = dose_mmol,
           treatment = "200 mg QD to steady state"),
    amt ~ time | treatment + id, doseu = "mmol")
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
    conc_obj, dose_obj,
    intervals = data.frame(start = t_last, end = t_last + tau, auclast = TRUE)))
  as.data.frame(res$result) |> filter(PPTESTCD == "auclast") |> pull(PPORRES)
}

# The nanocrystal-bound concentration decays through 40 orders of magnitude
# within one dosing interval and its far tail sits in solver round-off, where a
# value can go very slightly negative. PKNCA's log-down trapezoid takes log() of
# it and returns NaN, so clamp the round-off to zero first.
sim_ss$Cc_np   <- pmax(sim_ss$Cc_np, 0)
sim_ss$Cc_free <- sim_ss$Cc - sim_ss$Cc_np
p <- sim_ss[1, ]

identities <- tibble(
  Quantity = c("AUCtau(itraconazole, dissolved)",
               "AUCtau(itraconazole, nanocrystal-bound)",
               "AUCtau(hydroxy-itraconazole)"),
  `Closed form`  = c("Dose / CL_P", "Dose / (kN * V_P1)", "Dose / CL_M"),
  Predicted = c(dose_mmol / p$cl,
                dose_mmol / (p$kdiss * p$vc),
                dose_mmol / p$cl_ohi),
  `PKNCA AUCtau` = c(auc_ss("Cc_free"), auc_ss("Cc_np"), auc_ss("Cc_ohi"))
) |>
  mutate(Ratio = `PKNCA AUCtau` / Predicted)

identities |>
  knitr::kable(digits = c(0, 0, 5, 5, 5),
               caption = "Model-internal mass-balance identities at true steady state (200 mg QD, 70 kg, mmol*h/L).")
```

```{r ss-identity-checks}
# Deterministic: the only error sources are the ODE solver tolerance, the
# trapezoidal AUC rule on a 0.05 h grid, and the residual approach to steady
# state after 120 daily doses. 0.5% is generous for all three together and
# still catches any structural error (a missing elimination path, a wrong
# scaling volume, or an fm that is not 1 move these by tens of percent).
stopifnot(max(abs(identities$Ratio - 1)) < 0.005)
```

# Therapeutic drug monitoring targets

Jansen 2023 simulated day-14 trough concentrations for the standard 200 mg
once-daily NCF regimen and compared them against the itraconazole targets of
> 0.5 mg/L (prophylaxis) and > 1.0 mg/L (treatment), the itraconazole plus
hydroxy-itraconazole targets of > 1.0 mg/L (prophylaxis) and > 2.0 mg/L
(treatment), and the itraconazole concentration of > 4.0 mg/L associated with
toxicity. The published attainment was 100% / 54.4% for itraconazole, 100% /
91.8% for the sum, and none above the toxicity threshold. This is a partial
replication: the published Monte Carlo used the demographics of 1576
haematology patients from the authors' department, which are not published, so
the cohort here uses the trial's own weight distribution instead.

```{r targets}
day14 <- sim |>
  filter(time == 312) |>
  transmute(id,
            ITZ = Cc * MW_ITZ,
            SUM = Cc * MW_ITZ + Cc_ohi * MW_OHI)

targets <- tibble(
  Target = c("Itraconazole Cmin > 0.5 mg/L (prophylaxis)",
             "Itraconazole Cmin > 1.0 mg/L (treatment)",
             "Itraconazole + hydroxy Cmin > 1.0 mg/L (prophylaxis)",
             "Itraconazole + hydroxy Cmin > 2.0 mg/L (treatment)",
             "Itraconazole Cmin > 4.0 mg/L (toxicity)"),
  Published = c(100, 54.4, 100, 91.8, 0),
  Simulated = c(100 * mean(day14$ITZ > 0.5),
                100 * mean(day14$ITZ > 1.0),
                100 * mean(day14$SUM > 1.0),
                100 * mean(day14$SUM > 2.0),
                100 * mean(day14$ITZ > 4.0))
)

targets |>
  rename("Published (%)" = Published, "Simulated (%)" = Simulated) |>
  knitr::kable(digits = 1,
               caption = "Day-14 trough target attainment. The published column comes from a Monte Carlo over 1576 haematology patients whose demographics are not reported; the simulated column uses the trial's own weight distribution.")
```

```{r target-checks}
# Only the claims the paper states as absolutes are gated, and as proportions
# with headroom rather than as exact 0 / 100 (see pattern 12 of
# known-vignette-failure-patterns.md). The two treatment-target percentages are
# NOT gated: they depend on a covariate distribution the paper does not report.
stopifnot(
  mean(day14$ITZ > 0.5) > 0.95,
  mean(day14$SUM > 1.0) > 0.95,
  mean(day14$ITZ > 4.0) < 0.05
)
```

The itraconazole treatment target is the one row that does not reproduce, and
the reason is visible in the weight sensitivity below. The day-14 trough sits
almost exactly on the 1.0 mg/L threshold, so the percentage above it is set by
where the simulated population's weight distribution is centred - the one input
this replication cannot reproduce, because the demographics of the paper's
1576-patient database are not published. A cohort centred near 70 kg puts the
typical trough at 1.02 mg/L and therefore roughly half the population above the
threshold, which is what the published 54.4% implies; the trial's own median of
78.3 kg puts it at 0.92 mg/L and only 23% above.

```{r weight-sensitivity}
tibble(WT = c(70, 75, 78.3)) |>
  rowwise() |>
  mutate(res = list(rxode2::rxSolve(
    mod, add_obs(trial_doses, 312) |> mutate(WT = WT),
    omega = NA, addDosing = FALSE, returnType = "data.frame"))) |>
  mutate(`Itraconazole (mg/L)` = res$Cc[1] * MW_ITZ,
         `Itraconazole + hydroxy (mg/L)` = res$Cc[1] * MW_ITZ + res$Cc_ohi[1] * MW_OHI) |>
  ungroup() |>
  select(-res) |>
  rename("Body weight (kg)" = WT) |>
  knitr::kable(digits = 3,
               caption = "Typical-value day-14 trough as a function of body weight, showing the sensitivity of the 1.0 mg/L treatment target to the simulated population's weight distribution.")
```

# Assumptions and deviations

* **Correlated residual error is not reproduced.** The published `$SIGMA` is a
  `BLOCK(2)` with off-diagonal 0.0185, i.e. a correlation of 0.211 between the
  itraconazole and hydroxy-itraconazole log-scale residuals. Table 2 reports
  this as 48.5%, which is the same `sqrt(exp(x) - 1)` back-transformation the
  paper applies to its variance terms, evaluated at 0.211 - the two printings
  are consistent once that is recognised. nlmixr2 has no syntax for a
  correlated residual-error block across endpoints, so the two residuals are
  encoded independently. This affects the width of a joint predictive interval
  for the two analytes, not the typical-value predictions or either marginal
  residual magnitude.
* **Zero-variance IIV terms are omitted.** The control stream declares `$OMEGA`
  entries for `V_P1`/`V_N`, `Q_P`, `V_P2` and `V_M` and fixes all four to
  `0 FIX`. Carrying them would make the OMEGA matrix singular and break
  `rxSolve`'s Cholesky sampler, so only the two estimated IIV terms (on `CL_P`
  and `CL_M`) are declared. The model is unchanged.
* **Molar units, and the molecular weights used here.** The packaged model is
  in mmol and mmol/L because the source data were converted to molar
  equivalents before fitting (Methods) and Figure 1 is plotted in mmol/L. The
  molecular weights used in this vignette to convert to mg/L (705.6 g/mol for
  itraconazole, 721.6 g/mol for hydroxy-itraconazole) are **chemical constants
  from PubChem, not values printed in the paper**.
* **Body-weight distribution of the virtual cohort, and the one target that does
  not reproduce.** The paper reports only the median (78.3 kg) and range
  (60.0-92.5 kg) of the trial's weights, and does not report the demographics of
  the 1576-patient database used for its own Monte Carlo. The cohort here samples
  a log-normal with median 78.3 kg and a log-SD of 0.12, which puts the central
  95% of the distribution over roughly 62-99 kg. Against that cohort the
  itraconazole treatment target (Cmin > 1.0 mg/L) is met by about 23% of subjects
  versus the published 54.4%. This is a known, documented deviation rather than a
  transcription problem: the day-14 trough is 1.02 mg/L at 70 kg and 0.92 mg/L at
  78.3 kg, so the published 54.4% is what a cohort centred near 70 kg would give.
  Every other published attainment figure reproduces. The two treatment-target
  rows are deliberately excluded from the vignette's assertions for this reason.
* **`V_N = V_P1` is the paper's own assumption**, stated in the Discussion:
  nanocrystal-bound concentrations were never measured, so the volume of the
  nanocrystal compartment is not identifiable and was set equal to the
  itraconazole central volume.
* **`fm = 1` is the paper's own assumption**, made in the absence of metabolic
  conversion data. Every hydroxy-itraconazole parameter is therefore apparent
  to the fraction metabolised.
* **The published trough summary is compared against individual predictions.**
  The paper's mean Cmin values are arithmetic means of observed concentrations
  and so include residual error; the typical-value comparison above is
  noise-free. The log-scale residual would inflate a simulated arithmetic mean
  by about 11% (itraconazole) and 2% (hydroxy-itraconazole).
* **No erratum or corrigendum was found** for this article at the time of
  extraction.
