---
title: "Pyrazinamide (Jayanti 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pyrazinamide (Jayanti 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Jayanti_2026_pyrazinamide"))$reference`
* Description: `r rxode2::rxode(readModelDb("Jayanti_2026_pyrazinamide"))$description`
* Article: <https://doi.org/10.1371/journal.pone.0340133>
* Correction (funding statement only): <https://doi.org/10.1371/journal.pone.0347490>
* Supporting information (S1-S4 Tables, S1-S5 Figs, and the deposited NONMEM
  dataset as S2 File): <https://doi.org/10.1371/journal.pone.0340133#sec022>

Jayanti 2026 is the interethnic successor to the same cPMTb group's Korean-only
pyrazinamide (PZA) model, which this library also ships as
[`Kim_2023_pyrazinamide`](Kim_2023_pyrazinamide.html). It pools 160 Korean and
160 Indonesian tuberculosis patients, matched 1:1 on body weight and age, to ask
two questions: does ethnicity itself change PZA disposition, and does diabetes
mellitus (DM)?

The answers drive the unusual shape of the model. Ethnicity is **not** a
covariate effect -- "Ethnicity did not significantly affect the CL/F and Vd/F of
PZA" -- yet apparent clearance is nonetheless estimated *separately per
ethnicity*, because the two countries' diabetic patients differ in age: Indonesian
diabetic patients tend to be under 60, Korean diabetic patients tend to be older.
A single pooled diabetes term could not describe both, so the final model carries
two clearance typical values, two clearance variances, and two different diabetes
covariate forms, selected by ethnicity.

# Population

The cPMTb cohort is a multinational prospective tuberculosis cohort spanning 22
hospitals in the Republic of Korea and one in Surabaya, Indonesia, recruiting
between 2018-08-20 and 2021-11-03. Adults (>= 18 years) with drug-susceptible
tuberculosis on a PZA-based regimen for at least two weeks were eligible;
non-adherent, pregnant and multi-drug-resistant patients were excluded.

To remove demographic confounding from the interethnic comparison the authors
matched patients between countries on body weight and age (maximum difference
5 kg and 5 years), using the smaller Indonesian sample as the reference. That
gives the modelled cohort of **160 patients per ethnicity** contributing **407
concentrations** (240 Indonesian, 167 Korean -- Indonesian inpatients generally
contributed two samples each, Korean outpatients one).

Baseline characteristics (Table 1): median age 46 years (IQR 31-57), total body
weight 55 kg (IQR 50-59.8), lean body weight 45.5 kg (IQR 40.8-49.3), height
165 cm (IQR 160-170), 38.4% female, median eGFR 105.5 mL/min/1.73 m^2. Seventy-
seven patients (24.1%) had diabetes mellitus -- 55 of 160 Indonesians against 22
of 160 Koreans -- and 25 (7.8%) were both diabetic and aged 60 years or older.

Sampling was sparse and randomised: blood was drawn at a random time 0-24 h after
the last dose. PZA was quantified by validated HPLC-ESI-MS/MS over 2.0-80.0 mg/L
(LLOQ 2.0 mg/L). 117 below-LLOQ samples and 28 outlier patients were excluded;
the authors report refitting the base model with the below-LLOQ records included
and finding worse goodness-of-fit and poorer parameter precision.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Jayanti_2026_pyrazinamide")()$population`).

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Jayanti_2026_pyrazinamide.R` carries an in-file
comment naming its origin. They are collected here for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Structural model (1-cmt, first-order absorption and elimination) | n/a | Results, "Population pharmacokinetics model of PZA"; absorption-delay models all rejected |
| `lclIndonesian` (CL/F, Indonesian) | 3.18 L/h | Table 2, `CL/F Indonesian; = theta1` (RSE 4.7%) |
| `lclKorean` (CL/F, Korean) | 3.5 L/h | Table 2, `CL/F Korean; = theta2` (RSE 4.5%) |
| `lvc` (Vd/F) | 52.8 L | Table 2, `Vd/F (L) = theta5` (RSE 6.6%) |
| `lka` (Ka) | 2.0 1/h | Table 2, `Ka (h-1) = theta6` (RSE 15%) |
| `e_lbm_cl` | 0.75 (fixed) | Methods, "fixed exponents of 0.75 for CL/F and 1 for Vd/F" |
| `e_lbm_vc` | 1.0 (fixed) | Methods, same sentence |
| Lean-body-weight reference | **45 kg** | **Table 2 footnote d** (see below) |
| `e_diab_cl` (Indonesian DM) | 0.23 | Table 2, `CL/F; DM = theta1 x (1 + theta3)` (RSE 41.6%) |
| `e_age_ge60_diab_cl` (Korean old DM) | 0.26 | Table 2, `CL/F; Old DM = theta2 x (1 + theta4)` (RSE 42.3%) |
| Retained covariate combination | DM in Indonesians, old DM in Koreans | Results, "we included DM (in Indonesians) - OldDM (in Koreans) as covariates of CL/F in the final model"; S1 Table row 1 |
| Geriatric-diabetes indicator | `AGE_GE60 * DIS_DIAB` | Deposited dataset (S2 File) column `OLDDM`, which equals `OLD * DM` exactly |
| `etalclIndonesian` | 0.205 | Table 2, `omega^2; CL/F Indonesian (%)` = 20.5 |
| `etalclKorean` | 0.05 | Table 2, `omega^2; CL/F Korean (%)` = 5 |
| `etalvc` | 0.236 | Table 2, `omega^2; Vd/F (%)` = 23.6 |
| IIV on Ka | fixed to 0 (no eta) | Table 2, `omega^2; Ka (%)` = `0 (FIX)` |
| `addSd` | 1.29 mg/L | Table 2, `Residual variability / Additive` (RSE 14.6%) |

## The 45 kg allometric reference is only in the table image

The lean-body-weight value that every typical value in Table 2 refers to appears
**only in footnote d of Table 2**:

> Allometric scaling was applied to the CL/F and Vd/F data, and typical values
> reported here refer to the typical patient, with lean body weight of 45 kg.

That footnote is easy to lose. It survives in the publisher's rendered table
image (`pone.0340133.t002`) and in a layout-preserving text extraction of the
PDF, but it is dropped entirely by PDF-to-markdown conversion, which is the form
most readers of this article will work from. Without it the reference has to be
inferred, and the inference is ambiguous: back-solving the
per-subgroup post-hoc CL/F medians of S2 Table brackets it only to roughly
45-46 kg. The rounded population median is 45.5 kg in Table 1 and 45.56 kg
recomputed from the deposited dataset, so 45 kg is the rounded median as the
footnote implies.

## Reading the IIV scale

Table 2 labels its variability rows `omega^2; <parameter> (%)`, which on its own
could mean a variance, a CV%, or a variance expressed as a percentage. Three
independent lines of evidence agree that it is the **variance multiplied by
100**, so that `20.5` means `omega^2 = 0.205`:

1. **Table 2's own footnote a** defines the symbol: "`omega^2`, variance of
   interindividual variability".
2. **The same group uses the identical convention** in Kim 2023, where it is
   pinned by a row printed as `3 (FIX)` against the literal `(0.03 FIX)` of that
   paper's deposited NONMEM control stream.
3. **It is the only reading compatible with this paper's own post-hoc spread.**
   S2 Table gives the Korean non-old-DM CL/F interquartile range as
   3.22-3.91 L/h about a 3.47 L/h median. Removing the lean-body-weight
   contribution leaves an empirical-Bayes log-scale spread of about 0.093, which
   at the reported 56.9% shrinkage implies `omega` near 0.216 -- against 0.224
   predicted by `omega^2 = 0.05`, and 0.05 predicted by the CV% reading. The
   Indonesian arm points the same way.

```{r iiv-scale}
# Line 3 above, evaluated. sd(eta_EBE) = omega * (1 - shrinkage), and the
# interquartile range of a normal spans 2 * qnorm(0.75) = 1.349 standard
# deviations.
iqr_to_sd <- function(q1, q3) log(q3 / q1) / (2 * qnorm(0.75))

# Korean, not old-DM: S2 Table CL/F IQR 3.22-3.91; Table 2 shrinkage 56.9%.
# The lean-body-weight term contributes 0.75 * sdlog(LBW) to the CL/F spread;
# Table 1 gives the Korean LBW IQR as 41.6-50.75 kg.
sd_ebe_kor  <- iqr_to_sd(3.22, 3.91)
sd_lbw_kor  <- 0.75 * iqr_to_sd(41.6, 50.75)
omega_kor   <- sqrt(max(sd_ebe_kor^2 - sd_lbw_kor^2, 0)) / (1 - 0.569)

tibble::tibble(
  Reading            = c("variance x 100 (used here)", "CV percent"),
  `Implied omega`    = c(sqrt(0.05), 0.05),
  `omega from S2 Table` = omega_kor
) |>
  knitr::kable(digits = 3, caption = "Korean clearance: omega implied by each reading of Table 2, against omega recovered from the paper's own post-hoc interquartile range.")

# The CV% reading is not merely a worse fit, it is arithmetically impossible for
# the Indonesian arm: shrinkage can only make the post-hoc spread SMALLER than
# omega, yet the Indonesian non-DM post-hoc spread already exceeds the omega
# that reading would imply.
sd_ebe_ind <- iqr_to_sd(2.36, 3.8)   # S2 Table, Indonesian non-DM CL/F IQR
stopifnot(sd_ebe_ind > 0.205)        # > omega under the CV% reading (0.205)
stopifnot(sd_ebe_ind < sqrt(0.205))  # < omega under the variance reading (0.453)
```

# Virtual cohort

Original observed data are not publicly available in a per-patient form suitable
for re-simulation here, but the deposited NONMEM dataset (S2 File) *is* public
and is used below only to check covariate coding, never as a parameter source.

The cohort mirrors the four subgroups Jayanti 2026 uses for its post-hoc exposure
comparison in S2 Table and Fig 3: Indonesian with and without DM, and Korean
split into "old DM" (>= 60 years old and diabetic) and "other patients". Doses
are 1,200 mg once daily, the normalising dose of S2 Table and Fig 3.

Lean body weight is drawn per ethnicity from a log-normal centred on the Table 1
medians (Korean 46.2 kg, Indonesian 44.9 kg) with a log-scale spread recovered
from the Table 1 interquartile ranges, truncated to the range observed in the
deposited dataset. The paper does not report lean body weight separately by
diabetes stratum, so both strata of an ethnicity share a distribution.

```{r cohort}
rxode2::rxSetSeed(20260129)
set.seed(20260129)

dose_mg   <- 1200   # S2 Table and Fig 3 normalise exposures to 1,200 mg
tau       <- 24     # once daily
n_dose    <- 7      # 7 daily doses; t1/2 is about 10 h, so this is steady state
t_ss      <- tau * (n_dose - 1)
n_per_arm <- 200    # the per-arm cap

# Table 1 lean-body-weight medians and interquartile ranges, by ethnicity.
lbm_med <- c(Indonesian = 44.9,  Korean = 46.2)
lbm_sd  <- c(Indonesian = log(48.00 / 40.70) / (2 * qnorm(0.75)),
             Korean     = log(50.75 / 41.60) / (2 * qnorm(0.75)))
lbm_lo  <- 32.45    # deposited dataset (S2 File) observed range
lbm_hi  <- 68.17

# Observation grid: dense through the absorption phase (Tmax is near 1.6 h with
# Ka = 2.0 1/h) so the trapezoidal AUC does not understate the peak.
obs_grid <- sort(unique(c(seq(0, 4, by = 0.05), seq(4, tau, by = 0.25))))

make_arm <- function(n, arm, ethnicity, race_korean, dis_diab, age_ge60,
                     id_offset, sd_scale = 1) {
  subj <- tibble(
    id          = id_offset + seq_len(n),
    arm         = arm,
    RACE_KOREAN = race_korean,
    DIS_DIAB    = dis_diab,
    AGE_GE60    = age_ge60,
    LBM         = pmin(pmax(
      lbm_med[[ethnicity]] * exp(rnorm(n, 0, lbm_sd[[ethnicity]] * sd_scale)),
      lbm_lo), lbm_hi)
  )
  doses <- subj |>
    crossing(time = seq(0, tau * (n_dose - 1), by = tau)) |>
    mutate(amt = dose_mg, evid = 1L, cmt = "depot")
  obs <- subj |>
    crossing(time = t_ss + obs_grid) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}

events <- bind_rows(
  make_arm(n_per_arm, "Indonesian, DM",     "Indonesian", 0, 1, 0, id_offset =   0L),
  make_arm(n_per_arm, "Indonesian, non-DM", "Indonesian", 0, 0, 0, id_offset = 200L),
  make_arm(n_per_arm, "Korean, old DM",     "Korean",     1, 1, 1, id_offset = 400L),
  make_arm(n_per_arm, "Korean, other",      "Korean",     1, 0, 0, id_offset = 600L)
)

# Disjoint IDs across arms are mandatory -- duplicate IDs silently merge.
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))

events |> count(arm, name = "rows")
```

# Typical-value check

A `zeroRe()`-style solve with every subject pinned at the model's 45 kg
lean-body-weight reference must return Table 2's thetas exactly. These are
deterministic quantities, so they are asserted at machine tolerance rather than
with a cohort-sized band.

```{r typical}
mod <- readModelDb("Jayanti_2026_pyrazinamide")

typ_events <- bind_rows(
  make_arm(1, "Indonesian, DM",     "Indonesian", 0, 1, 0, id_offset = 1000L, sd_scale = 0),
  make_arm(1, "Indonesian, non-DM", "Indonesian", 0, 0, 0, id_offset = 1001L, sd_scale = 0),
  make_arm(1, "Korean, old DM",     "Korean",     1, 1, 1, id_offset = 1002L, sd_scale = 0),
  make_arm(1, "Korean, other",      "Korean",     1, 0, 0, id_offset = 1003L, sd_scale = 0),
  make_arm(1, "Korean, young DM",   "Korean",     1, 1, 0, id_offset = 1004L, sd_scale = 0)
) |>
  mutate(LBM = 45)   # pin at the Table 2 footnote d reference

sim_typ <- rxode2::rxSolve(
  mod, events = typ_events, omega = NA, sigma = NA,
  keep = c("arm", "LBM", "RACE_KOREAN", "DIS_DIAB", "AGE_GE60")
) |>
  as.data.frame() |>
  mutate(tad = time - t_ss)

typ_par <- sim_typ |>
  group_by(arm) |>
  summarise(cl = first(cl), vc = first(vc), ka = first(ka), .groups = "drop")

# Table 2 arithmetic, written out independently of the model.
expected <- tibble::tribble(
  ~arm,                  ~cl_expected,
  "Indonesian, DM",      3.18 * (1 + 0.23),
  "Indonesian, non-DM",  3.18,
  "Korean, old DM",      3.50 * (1 + 0.26),
  "Korean, other",       3.50,
  "Korean, young DM",    3.50   # DM alone is NOT a Korean covariate
)

chk_typ <- typ_par |> left_join(expected, by = "arm")
stopifnot(
  nrow(chk_typ) == 5L,
  !anyNA(chk_typ$cl_expected),
  max(abs(chk_typ$cl - chk_typ$cl_expected)) < 1e-8,
  max(abs(chk_typ$vc - 52.8)) < 1e-8,
  max(abs(chk_typ$ka - 2.0)) < 1e-8
)

chk_typ |>
  rename("Subgroup" = arm, "CL/F (L/h)" = cl, "Vd/F (L)" = vc,
         "Ka (1/h)" = ka, "CL/F expected from Table 2 (L/h)" = cl_expected) |>
  knitr::kable(digits = 4, caption = "Typical-value parameters at the 45 kg lean-body-weight reference of Table 2 footnote d. Table 2 gives CL/F = 3.18 (Indonesian) and 3.5 (Korean) L/h, Vd/F = 52.8 L and Ka = 2.0 1/h; the Results text quotes 3.88 and 4.38 L/h for Indonesian and older Korean diabetic patients, against 3.91 and 4.41 here (the table's values are rounded to two significant figures).")
```

Note the fifth row. A **Korean** diabetic patient younger than 60 takes the
Korean reference clearance unmodified, because diabetes enters the Korean arm
only crossed with age; this is exactly what the Results mean by "3.5 L/h for
younger Korean patients with DM". An **Indonesian** diabetic patient takes the
23% increase at any age.

# Steady-state mass balance

At steady state under linear one-compartment kinetics, AUC over one dosing
interval equals `Dose / (CL/F)` exactly, whatever the absorption model. Both
sides use the same drawn parameters, so the only difference is trapezoidal
error and the bound can be tight.

```{r mass-balance}
mb <- sim_typ |>
  filter(!is.na(Cc), tad >= 0) |>
  group_by(arm) |>
  summarise(
    auc_trapz = sum(diff(tad) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    auc_ident = dose_mg / first(cl),
    .groups   = "drop"
  ) |>
  mutate(pct_diff = 100 * (auc_trapz - auc_ident) / auc_ident)

# Deterministic: both sides use the same drawn parameters, so the only error is
# trapezoidal. Realised max 0.005% on the grid above; 0.1% still goes red if the
# observation grid is ever coarsened enough to understate the peak.
stopifnot(nrow(mb) == 5L, max(abs(mb$pct_diff)) < 0.1)

mb |>
  rename("Subgroup" = arm, "AUC0-24 (trapezoid, mg*h/L)" = auc_trapz,
         "Dose / (CL/F) (mg*h/L)" = auc_ident, "Difference (%)" = pct_diff) |>
  knitr::kable(digits = c(0, 2, 2, 4), caption = "Steady-state AUC0-24 equals Dose/(CL/F) for the typical subject in every subgroup.")
```

This identity also holds in the paper's own numbers, which is worth recording
because it does **not** hold in the predecessor Kim 2023 (whose published AUC0-24
column is about half of `Dose/(CL/F)`). Every one of the six S2 Table subgroups
here reconciles to within 1.5%:

```{r s2-identity}
s2 <- tibble::tribble(
  ~group,                ~cl_pub, ~auc_pub,
  "Korean, old DM",         4.61,    261.4,
  "Korean, other",          3.47,    345.7,
  "Korean, total",          3.52,    343.2,
  "Indonesian, DM",         3.74,    322.2,
  "Indonesian, non-DM",     3.12,    388.6,
  "Indonesian, total",      3.23,    371.1
) |>
  mutate(auc_from_cl = 1200 / cl_pub,
         pct_diff    = 100 * (auc_pub - auc_from_cl) / auc_from_cl)

# Deterministic -- both columns are transcribed published constants. Realised
# max 1.04%, consistent with the tables' 3-4 significant figures.
stopifnot(max(abs(s2$pct_diff)) < 1.5)

s2 |>
  rename("S2 Table subgroup" = group, "Published CL/F (L/h)" = cl_pub,
         "Published AUC0-24 (mg*h/L)" = auc_pub,
         "1200 / CL/F (mg*h/L)" = auc_from_cl, "Difference (%)" = pct_diff) |>
  knitr::kable(digits = c(0, 2, 1, 1, 2), caption = "The published S2 Table is internally consistent: its AUC0-24 column reproduces Dose/(CL/F) from its own CL/F column.")
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod, events = events,
  keep = c("arm", "LBM", "RACE_KOREAN", "DIS_DIAB", "AGE_GE60")
) |>
  as.data.frame() |>
  mutate(tad = time - t_ss)

stopifnot(nrow(sim) > 0, !all(is.na(sim$Cc)), all(sim$Cc[!is.na(sim$Cc)] >= 0))
```

# Replicate published figures

```{r figure-2}
# Analogous to Fig 2 of Jayanti 2026 (prediction-corrected VPC): the spread of
# PZA concentrations across one steady-state dosing interval. The paper plots
# the observed median inside a 90% prediction interval.
sim |>
  filter(!is.na(Cc)) |>
  group_by(arm, tad) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = quantile(Cc, 0.50),
            Q95 = quantile(Cc, 0.95), .groups = "drop") |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  facet_wrap(~arm) +
  labs(x = "Time after dose (h)", y = "PZA concentration (mg/L)",
       title = "Steady-state PZA concentrations, 1,200 mg once daily",
       caption = paste("Median with 5th-95th percentile band,", n_per_arm,
                       "simulated subjects per arm. Analogous to Fig 2 of Jayanti 2026."))
```

```{r figure-3}
# Replicates Fig 3 of Jayanti 2026: AUC0-24, Cmax and CL/F contrasted between
# the DM and non-DM strata of each ethnicity.
per_subject <- sim |>
  filter(!is.na(Cc), tad >= 0) |>
  group_by(arm, id) |>
  summarise(
    Cmax   = max(Cc),
    `CL/F` = first(cl),
    `Vd/F` = first(vc),
    AUC024 = sum(diff(tad) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    .groups = "drop"
  )

per_subject |>
  pivot_longer(c(AUC024, Cmax, `CL/F`), names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric,
    AUC024 = "AUC0-24 (mg*h/L)", Cmax = "Cmax (mg/L)", `CL/F` = "CL/F (L/h)")) |>
  ggplot(aes(arm, value, fill = arm)) +
  geom_boxplot(outlier.size = 0.4, show.legend = FALSE) +
  facet_wrap(~metric, scales = "free_y") +
  labs(x = NULL, y = NULL,
       title = "Exposure and clearance by ethnicity and diabetes stratum (1,200 mg daily)",
       caption = "Replicates Fig 3 of Jayanti 2026.") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

The model reproduces the paper's qualitative ordering: within each ethnicity the
diabetic stratum has the higher CL/F and the lower exposure, and the Korean
old-DM arm has the lowest exposure of the four -- "Compared to Indonesian TB-DM
patients, older DM patients among Korean subjects had the lowest exposure of
PZA".

# PKNCA validation

Steady-state NCA over the final dosing interval. The NCA runs on `Cc`, the
individual prediction, rather than on `sim`: S2 Table's values are post-hoc
model-derived quantities with no residual error in them, and `Cmax` taken from a
residual-error-carrying simulation is biased upward.

```{r pknca}
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, arm)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id,
                             concu = "mg/L", timeu = "h")

dose_df <- events |>
  filter(evid == 1) |>
  select(id, time, amt, arm)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id,
                             route = "extravascular", doseu = "mg")

intervals <- data.frame(
  start = t_ss, end = t_ss + tau,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, cav = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r pknca-summary}
as.data.frame(nca_res) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "cmin", "auclast", "cav")) |>
  group_by(arm, PPTESTCD) |>
  summarise(median = median(PPORRES), .groups = "drop") |>
  pivot_wider(names_from = PPTESTCD, values_from = median) |>
  rename("Subgroup" = arm, "AUC0-24 (mg*h/L)" = auclast, "Cavg (mg/L)" = cav,
         "Cmax (mg/L)" = cmax, "Cmin (mg/L)" = cmin, "Tmax (h)" = tmax) |>
  knitr::kable(digits = 2, caption = paste0("Simulated steady-state NCA (medians of ", n_per_arm, " subjects per arm), 1,200 mg once daily."))
```

## Comparison against published NCA

S2 Table reports median post-hoc Cmax and AUC0-24, normalised to 1,200 mg, for
exactly these four subgroups.

```{r nca_compare}
published <- tibble::tribble(
  ~arm,                  ~cmax,  ~auclast,
  "Indonesian, DM",       25.27,    322.2,
  "Indonesian, non-DM",   28.70,    388.6,
  "Korean, old DM",       21.46,    261.4,
  "Korean, other",        25.83,    345.7
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "arm",
  params        = c("cmax", "auclast"),
  units         = c(cmax = "mg/L", auclast = "mg*h/L"),
  tolerance_pct = 20
)

knitr::kable(cmp,
  caption = "Simulated vs. published (S2 Table) steady-state NCA at 1,200 mg once daily. * differs from reference by >20%.",
  align   = c("l", "l", "r", "r", "r"))
```

```{r nca_gate}
# Cohort-derived quantities, so the bound is on the CENTRE and is set with
# headroom over the sampling noise, not tightened to one run. With 200 subjects
# per arm the median of a log-normal with the Indonesian omega of 0.45 carries a
# standard error near 4%, and the systematic subgroup differences below run
# 2-8%. A 20% bound still goes red on a mis-transcribed clearance, dose, volume
# or unit, each of which moves these numbers by tens of percent.
sim_med <- as.data.frame(nca_res) |>
  filter(PPTESTCD %in% c("cmax", "auclast")) |>
  group_by(arm, PPTESTCD) |>
  summarise(sim = median(PPORRES), .groups = "drop") |>
  pivot_wider(names_from = PPTESTCD, values_from = sim)

gate <- published |>
  rename(cmax_pub = cmax, auclast_pub = auclast) |>
  left_join(sim_med, by = "arm") |>
  mutate(cmax_pct = 100 * (cmax - cmax_pub) / cmax_pub,
         auc_pct  = 100 * (auclast - auclast_pub) / auclast_pub)

stopifnot(
  nrow(gate) == 4L,                      # guard against a silent zero-row join
  !anyNA(gate$cmax), !anyNA(gate$auclast),
  max(abs(gate$cmax_pct)) < 20,
  max(abs(gate$auc_pct))  < 20
)

gate |>
  select(arm, cmax_pct, auc_pct) |>
  rename("Subgroup" = arm, "Cmax difference (%)" = cmax_pct,
         "AUC0-24 difference (%)" = auc_pct) |>
  knitr::kable(digits = 1, caption = "Simulated minus published, as a percentage of published.")
```

Apparent volume of distribution agrees too, though with visibly more scatter than
clearance -- unsurprisingly, since `Vd/F` carries the model's largest variance
(`omega^2 = 0.236`) and is the parameter that randomly-timed single samples
identify worst (24.6% shrinkage).

```{r vd-compare}
vd_pub <- c("Indonesian, DM" = 51.27, "Indonesian, non-DM" = 48.29,
            "Korean, old DM" = 56.72, "Korean, other" = 50.60)

vd_cmp <- per_subject |>
  group_by(arm) |>
  summarise(vd_sim = median(`Vd/F`), .groups = "drop") |>
  mutate(vd_pub = vd_pub[arm],
         pct    = 100 * (vd_sim - vd_pub) / vd_pub)

# Cohort-derived, and noisier than the clearance-driven metrics: the median of
# 200 draws from a log-normal with total log-scale spread near 0.51 has a
# standard error of about 4.5%, so the realised -3 to +9% scatter is sampling
# noise about a correct centre rather than bias. 25% retains headroom over that
# while still going red on a mis-transcribed volume or allometric exponent.
stopifnot(nrow(vd_cmp) == 4L, !anyNA(vd_cmp$vd_pub), max(abs(vd_cmp$pct)) < 25)

vd_cmp |>
  rename("Subgroup" = arm, "Simulated Vd/F (L)" = vd_sim,
         "S2 Table Vd/F (L)" = vd_pub, "Difference (%)" = pct) |>
  knitr::kable(digits = 1, caption = "Apparent volume of distribution, simulated median against S2 Table.")
```

# Probability of target attainment

Jayanti 2026's dose recommendations rest on the probability of attaining
`AUC0-24 >= 363 mg*h/L`. Because steady-state `AUC0-24` is exactly `Dose/(CL/F)`
-- the identity gated above -- the attainment probability can be read straight
off the model's own simulated clearances without a second ODE solve.

Lean body weight per WHO weight band is taken from the deposited dataset (S2
File) rather than reconstructed from the paper's virtual-population body weights,
because the Boer equation the paper uses to derive lean body weight extrapolates
badly at low body weight: applied to the paper's own `<40 kg` virtual population
(34.75 kg) at the cohort median height it returns a lean body weight of about
39 kg, i.e. **greater than total body weight**. That pathology is real in the
deposited data too -- 12 of its 300 patients have a lean body weight exceeding
their total body weight.

```{r pta-cohort}
bands <- tibble::tribble(
  ~band,       ~lbm_med, ~lbm_sdlog,
  "<40 kg",       36.84,     0.0644,
  "40-54 kg",     42.34,     0.1030,
  "55-70 kg",     49.26,     0.0734,
  ">70 kg",       57.10,     0.0299
)

groups <- tibble::tribble(
  ~arm,                  ~RACE_KOREAN, ~DIS_DIAB, ~AGE_GE60,
  "Indonesian, DM",                 0,         1,         0,
  "Indonesian, non-DM",             0,         0,         0,
  "Korean, old DM",                 1,         1,         1,
  "Korean, other",                  1,         0,         0
)

n_pta <- 200
pta_subj <- tidyr::crossing(bands, groups) |>
  mutate(k = row_number()) |>
  rowwise() |>
  do({
    r <- .
    tibble(id = (r$k - 1) * n_pta + seq_len(n_pta),
           band = r$band, arm = r$arm,
           RACE_KOREAN = r$RACE_KOREAN, DIS_DIAB = r$DIS_DIAB,
           AGE_GE60 = r$AGE_GE60,
           LBM = pmin(pmax(r$lbm_med * exp(rnorm(n_pta, 0, r$lbm_sdlog)),
                           lbm_lo), lbm_hi))
  }) |>
  ungroup()

pta_events <- bind_rows(
  pta_subj |> mutate(time = 0, amt = 1200, evid = 1L, cmt = "depot"),
  pta_subj |> mutate(time = 1, amt = NA_real_, evid = 0L, cmt = "central")
) |>
  arrange(id, time, desc(evid))

pta_cl <- rxode2::rxSolve(
  mod, events = pta_events, keep = c("band", "arm")
) |>
  as.data.frame() |>
  group_by(band, arm, id) |>
  summarise(cl = first(cl), .groups = "drop")

stopifnot(nrow(pta_cl) == nrow(bands) * nrow(groups) * n_pta)
```

```{r pta-curves, fig.height = 5.5}
target_auc <- 363    # mg*h/L, Jayanti 2026 Methods and Results
doses      <- c(800, 1000, 1200, 1250, 1500, 1600, 2000, 2500, 3000)

pta_model <- tidyr::crossing(pta_cl, dose = doses) |>
  group_by(band, arm, dose) |>
  summarise(pta = 100 * mean(dose / cl >= target_auc), .groups = "drop")

# Published S4 Table (optimal-dose simulation).
pta_pub <- tibble::tribble(
  ~band,     ~dose, ~`Indonesian, DM`, ~`Indonesian, non-DM`, ~`Korean, old DM`, ~`Korean, other`,
  "<40 kg",   1000,  92.1, 95.4, 89.0, 94.2,
  "<40 kg",   1250,  99.1, 99.6, 94.0, 99.4,
  "<40 kg",   1500,  99.8, 99.9, 99.9, 99.9,
  "40-54 kg", 1000,  81.2, 87.8, 76.2, 85.6,
  "40-54 kg", 1250,  96.9, 98.4, 94.3, 97.8,
  "40-54 kg", 1500,  99.5, 99.8, 99.4, 99.7,
  "55-70 kg", 1000,  63.7, 72.5, 57.0, 62.3,
  "55-70 kg", 1250,  86.4, 89.4, 83.0, 85.0,
  "55-70 kg", 1500,  95.8, 98.8, 94.1, 97.0,
  ">70 kg",   1000,  42.2, 52.0, 38.0, 48.4,
  ">70 kg",   1250,  79.0, 83.0, 73.0, 83.0,
  ">70 kg",   1500,  85.5, 88.5, 84.5, 89.5
) |>
  pivot_longer(-c(band, dose), names_to = "arm", values_to = "pta_pub")

pta_cmp <- pta_model |> inner_join(pta_pub, by = c("band", "arm", "dose"))
stopifnot(nrow(pta_cmp) == nrow(pta_pub))

# Fail loudly rather than returning a zero-length vector that would make any
# downstream comparison vacuously true.
pick <- function(b, a, d) {
  v <- pta_model$pta[pta_model$band == b & pta_model$arm == a & pta_model$dose == d]
  if (length(v) != 1L) stop("no unique PTA cell for ", b, " / ", a, " / ", d)
  v
}

ggplot(pta_model, aes(dose, pta, colour = band)) +
  geom_line(linewidth = 0.8) +
  geom_point(data = pta_cmp, aes(dose, pta_pub, colour = band), shape = 1, size = 2.4) +
  geom_hline(yintercept = 90, linetype = "dashed", colour = "grey40") +
  facet_wrap(~arm) +
  labs(x = "Daily pyrazinamide dose (mg)", y = "Probability of AUC0-24 >= 363 mg*h/L (%)",
       colour = "WHO weight band",
       title = "Probability of target attainment",
       caption = paste("Lines: this model. Open circles: published S4 Table.",
                       "Dashed line: the paper's 90% PTA criterion.",
                       "Replicates Fig 4 of Jayanti 2026."))
```

**The published PTA tables are not reproducible from the published model, and the
disagreement is in the source.** The model's attainment curves are markedly
flatter and lower than S3/S4 Table at every band:

```{r pta-deviation}
pta_cmp |>
  mutate(diff = pta - pta_pub) |>
  group_by(band) |>
  summarise(`Median model PTA (%)` = median(pta),
            `Median published PTA (%)` = median(pta_pub),
            `Median difference (points)` = median(diff),
            `Largest difference (points)` = diff[which.max(abs(diff))],
            .groups = "drop") |>
  rename("WHO weight band" = band) |>
  knitr::kable(digits = 1, caption = "Model minus published probability of target attainment, over the 12 band-by-dose cells S4 Table reports.")
```

The decisive check needs no model at all: **the paper's own S2 Table implies the
same attainment this model does, and both disagree with S3 Table.** S2 Table
reports the Indonesian non-DM post-hoc AUC0-24 at 1,200 mg as a median of
388.6 mg*h/L with an interquartile range of 318.1-515.4. Fitting a log-normal to
those three published numbers and reading off the fraction above the paper's own
363 mg*h/L target gives an attainment that lands on top of the model's, and
nowhere near the 97.8% S3 Table prints for that stratum and dose.

```{r pta-three-way}
# (a) This model, Indonesian non-DM, 40-54 kg band, 1,200 mg.
pta_model_cell <- pick("40-54 kg", "Indonesian, non-DM", 1200)

# (b) Implied by S2 Table's OWN post-hoc AUC0-24 distribution for the same
#     stratum at the same dose: log-normal through median 388.6 and IQR
#     318.1-515.4, evaluated at the paper's 363 mg*h/L target.
s2_med <- 388.6; s2_q1 <- 318.1; s2_q3 <- 515.4
s2_sdlog <- log(s2_q3 / s2_q1) / (2 * qnorm(0.75))
pta_from_s2 <- 100 * pnorm(log(s2_med / target_auc) / s2_sdlog)

# (c) As printed in S3 Table for the WHO 1,200 mg dose, 40-54 kg band.
pta_published <- 97.8

tibble::tibble(
  Source = c("This model", "Implied by S2 Table's own post-hoc AUC quartiles",
             "As printed in S3 Table"),
  `PTA (%)` = c(pta_model_cell, pta_from_s2, pta_published)
) |>
  knitr::kable(digits = 1, caption = "Probability of AUC0-24 >= 363 mg*h/L for Indonesian non-diabetic patients in the 40-54 kg band at 1,200 mg, computed three ways.")

# The model agrees with the paper's own post-hoc distribution far better than
# that distribution agrees with the paper's PTA table. This is the gate: it goes
# red if the model ever stops reproducing S2 Table, which is the claim being
# made -- it is NOT a gate on the S3 Table value, which is the documented
# deviation.
stopifnot(abs(pta_model_cell - pta_from_s2) < 15,
          abs(pta_from_s2 - pta_published) > 25)
```

Two further observations point the same way.

1. **The published PTA barely differs between ethnicities, but the model's
   variances differ by a factor of four.** Table 2 gives the Indonesian clearance
   variance as 0.205 and the Korean as 0.05, a factor of two in `omega`. Two
   populations whose clearance spreads differ that much cannot produce the nearly
   superimposable attainment curves S3/S4 Table report -- the `<40 kg` band at
   800 mg is 70.4% for *both* ethnicities, to the decimal.
2. **The slope of the published dose-response implies a much narrower spread
   than Table 2.** Reading `omega` off consecutive S4 Table doses in the
   40-54 kg band gives about 0.23, which matches the Korean `omega` (0.224) and
   not the Indonesian one (0.453).

Together these suggest the Monte Carlo behind S3/S4 Table did not use the final
model as Table 2 reports it -- a single shared `omega`, and possibly a different
allometric reference, would both push attainment up. No parameter was tuned to
close the gap; the model file follows Table 2, which is what S2 Table
corroborates.

The paper's *qualitative* dose ranking survives, and is what a user should take
from the PTA analysis:

```{r pta-claims}
claims <- tibble::tribble(
  ~Claim, ~Model,
  "WHO 800 mg under-doses the <40 kg band (paper: 70.4%)",     pick("<40 kg", "Indonesian, non-DM", 800),
  "Raising <40 kg to 1,250 mg improves attainment",            pick("<40 kg", "Indonesian, non-DM", 1250),
  "Korean old DM attains less than Korean other, <40 kg 1,250 mg",
    pick("<40 kg", "Korean, old DM", 1250),
  "Indonesian DM attains less than Indonesian non-DM, 40-54 kg 1,250 mg",
    pick("40-54 kg", "Indonesian, DM", 1250)
)

# Directional claims only: magnitudes are the documented deviation above.
stopifnot(
  pick("<40 kg", "Indonesian, non-DM", 1250) > pick("<40 kg", "Indonesian, non-DM", 800),
  pick("<40 kg", "Korean, old DM", 1250)     < pick("<40 kg", "Korean, other", 1250),
  pick("40-54 kg", "Indonesian, DM", 1250)   < pick("40-54 kg", "Indonesian, non-DM", 1250),
  pick(">70 kg", "Indonesian, non-DM", 1000) < pick("<40 kg", "Indonesian, non-DM", 1000)
)

claims |>
  rename("Model PTA (%)" = Model) |>
  knitr::kable(digits = 1, caption = "Directional claims from the paper's dose exploration, evaluated on the packaged model.")
```

# Assumptions and deviations

- **Lean-body-weight distribution.** Table 1 reports medians and interquartile
  ranges by ethnicity but not a distributional form. Lean body weight is drawn
  log-normally, centred on the Table 1 medians, with the log-scale spread
  recovered from the Table 1 interquartile ranges and truncated to the range
  observed in the deposited dataset (32.45-68.17 kg). The paper does not report
  lean body weight by diabetes stratum, so both strata of an ethnicity share a
  distribution. For the weight-band PTA analysis the per-band lean body weight is
  taken from the deposited dataset instead, for the Boer-extrapolation reason
  given in that section.
- **Post-hoc versus prior-predictive.** S2 Table and Fig 3 report medians of
  *individual Bayesian post-hoc* estimates, each informed by that patient's own
  concentrations and shrunk toward the typical value (35.9% and 56.9% shrinkage
  on the two clearance terms, 24.6% on volume). The simulation here is a
  prior-predictive draw from the population model, so it reproduces the centre of
  those distributions but is wider in the tails by construction.
- **Subgroup sizes.** Simulated arms are equal-sized (200 each). The real
  subgroups were very unequal -- 13 Korean old-DM patients against 147 others,
  55 Indonesian diabetic against 105 non-diabetic -- so the published medians for
  the small strata carry far wider uncertainty than the simulated ones.
- **Dose and steady state.** Doses are 1,200 mg once daily to match the
  normalising dose of S2 Table and Fig 3; the real cohort received 500-2,000 mg,
  most commonly 1,500 mg (42.8%). Seven daily doses are simulated and NCA is
  taken over the seventh interval, which is well past steady state for a ~10 h
  half-life. The deposited dataset codes every dose record as `SS = 1` with
  `II = 24`, confirming the paper fitted steady-state data.
- **IIV on Ka is omitted, not encoded as a zero-variance eta.** Table 2 reports
  `omega^2; Ka (%)` as `0 (FIX)`. Declaring `etalka ~ fixed(0)` would put a
  structural zero on the OMEGA diagonal and make it singular, which fails at the
  Cholesky factorisation during simulation. Omitting the eta is the faithful
  encoding of a variance fixed to zero.
- **Covariates screened but not retained** (total body weight, continuous age,
  height, sex, albumin, AST, ALT, total bilirubin, blood urea nitrogen, serum
  creatinine, eGFR, liver disease and ethnicity-as-a-covariate) are recorded in
  the model file's `covariatesDataExcluded` metadata rather than `covariateData`,
  since none is referenced in `model()`. Albumin and AST are the notable cases:
  both were significant univariately and were dropped for large standard errors.
- **All parameter values come from the paper's Table 2, including its footnote
  d.** No value was taken from author correspondence, figure digitisation, or an
  upstream model. The deposited dataset (S2 File) was used only to resolve
  covariate *coding* -- the age threshold, the ethnicity code, the sex code and
  the `OLDDM` definition -- and to characterise lean body weight per weight band
  for the PTA section; no parameter estimate came from it.

## Errata and internal inconsistencies in the source

- **Published PTA tables are not reproducible from the published model.** See the
  target-attainment section; S3/S4 Table imply a clearance spread near half of
  the Indonesian `omega` reported in Table 2 and are inconsistent with the
  paper's own S2 Table exposure quartiles. The model file follows Table 2.
- **The 45 kg allometric reference appears only in the Table 2 footnote**, which
  PDF-to-markdown conversion drops (a layout-preserving text extraction and the
  rendered table image both keep it). Recorded here and in the model file's `LBM`
  covariate notes so it is not re-derived.
- **The age threshold is printed inconsistently.** Table 1's row header and the
  S3/S4 Table footnotes say "Age > 60 years old"; Methods, Results and the Fig 3
  caption say ">= 60 years". The deposited dataset settles it -- the `OLD` column
  is 0 for ages 17-59 and 1 for ages 60-80 -- so the inclusive form is correct and
  the model uses `AGE_GE60`.
- **Vd/F and Ka are quoted differently in the Discussion than in Table 2.** The
  Discussion says "The Vd/F (53.4 L) and Ka (2.1 h-1) values of our model", while
  Table 2 gives 52.8 L and 2.0 1/h with bootstrap medians of 52.7 and 2.05. The
  model uses the Table 2 values, per the standing rule that the final parameter
  table takes precedence over prose.
- **The Abstract mislabels the two clearance typical values.** It reads "There
  were 23% and 26% increases in apparent clearance (CL/F) of Indonesian patients
  with DM (CL/F 3.18 L/h) and Korean patients aged > 60 years with DM (CL/F
  3.5)", attaching the *reference* (non-diabetic) clearances to the diabetic
  groups. The Results give the diabetic values correctly as 3.88 and 4.38 L/h.
- **The DM coefficients are printed to different precision in different places.**
  Table 2 gives 0.23 and 0.26; the Results text gives 0.226 and 0.252. The model
  uses the Table 2 values; the difference moves the diabetic clearances by under
  1%.
- **Lean body weight exceeds total body weight for 12 of the 300 deposited
  patients** (up to a ratio of 1.14), because the Boer equation is extrapolated
  outside its validity for low-BMI tuberculosis patients. This is a property of
  the covariate definition the paper adopted, not a transcription error, but it
  means lean body weight should not be reconstructed from independently sampled
  weight and height at the low end of the weight range.
- **The deposited dataset is slightly smaller than the modelled cohort** -- 300
  subjects and 404 quantifiable records against the 320 subjects and 407
  concentrations described in the Results. The parameter estimates come from the
  paper, so this affects only the covariate-coding checks above.
- **The published correction** (PLoS One 2026;21(4):e0347490) revises the funding
  statement only and changes no model value.
