---
title: "Methotrexate nanoparticles and nanoemulsions in rats (Jeong 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methotrexate nanoparticles and nanoemulsions in rats (Jeong 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
suppressPackageStartupMessages({
  library(nlmixr2lib)
  library(PKNCA)
  library(rxode2)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})
```

# Model and source

Jeong 2021 reanalysed the plasma data of two earlier rat formulation studies
from the same group (methotrexate-loaded PLGA nanoparticles, Jang 2019;
methotrexate-loaded olive-oil/Labrasol nanoemulsions, Jang 2020) with Phoenix
NLME. The paper reports two final population PK models, each fitted to its own
data set, and both are packaged here:

* `Jeong_2021_methotrexate_nanoformulation_rat` -- free methotrexate solution
  versus any nanoformulation (Section 3.2, Equations (1)-(5), Table 3), fitted
  to all 40 rats.
* `Jeong_2021_methotrexate_nanoemulsion_rat` -- PLGA nanoparticles versus
  nanoemulsions (Section 3.4, Equations (6)-(10), Table 9), fitted to the 20
  rats that received a nanoformulation.

* Citation: `r rxode2::rxode2(readModelDb("Jeong_2021_methotrexate_nanoemulsion_rat"))$reference`
* Article: Pharmaceutics 2021;13(7):1050. <https://doi.org/10.3390/pharmaceutics13071050>

Both models are one-compartment with first-order absorption and elimination,
oral bioavailability, exponential IIV and a log-additive residual error. The
formulation indicator enters every covariate relationship as a linear
deviation, `P = tvP * (1 + dPdFormulation * indicator) * exp(eta)`.

# Population

Male Sprague-Dawley rats, 7-9 weeks old and 240-260 g, without disease
(Jeong 2021 Section 2.1). Each rat received one single dose by one route, with
5 rats per arm:

| Arm | Oral dose | IV dose | Source study |
|-----|-----------|---------|--------------|
| Free methotrexate solution | 5 mg/kg | 5 mg/kg | PLGA nanoparticle study (Figure 1) |
| Free methotrexate solution | 0.06 mg/kg | 0.024 mg/kg | Nanoemulsion study (Figure 2) |
| PLGA nanoparticles | 5 mg/kg | 5 mg/kg | Table 6 |
| Olive-oil/Labrasol nanoemulsions | 0.06 mg/kg | 0.024 mg/kg | Table 6 |

The two models' `population` metadata carry the same information
(`readModelDb("Jeong_2021_methotrexate_nanoemulsion_rat")$population`).

# Source trace

| Equation / parameter | Free-vs-nano model | NP-vs-NE model | Source location |
|----------------------|--------------------|----------------|-----------------|
| Structure: 1-cmt, first-order absorption, IV into central | -- | -- | Section 3.2 and 3.4; Figure 7 |
| Formulation effects as linear deviation | Eq. (1), (2), (4), (5) | Eq. (6), (7), (9), (10) | Sections 3.2, 3.4 |
| `lvc` (tvV) | 14.889 L/kg | 18.832 L/kg | Table 3 / Table 9 |
| `lcl` (tvCL) | 14.577 L/h/kg | 9.167 L/h/kg | Table 3 / Table 9 |
| `lka` (tvKa) | 0.582 1/h | 0.714 1/h | Table 3 / Table 9 |
| `lfdepot` (tvF) | 0.272 | 0.334 | Table 3 / Table 9 |
| `e_form_*_vc` (dVdFormulation) | 0.429 | -0.986 | Table 3 / Table 9 |
| `e_form_*_cl` (dCLdFormulation) | -0.355 | -0.990 | Table 3 / Table 9 |
| `e_form_*_ka` (dKadFormulation) | 10.883 | 1.552 | Table 3 / Table 9 |
| `e_form_*_fdepot` (dFdFormulation) | 4.246 | 0.193 | Table 3 / Table 9 |
| `etalvc` | 6.3504e-6 (IIV 0.252%) | 5.6644e-6 (IIV 0.238%) | Table 3 / Table 9 IIV (%) column |
| `etalcl` | 0.338 | 0.0084622 (IIV 9.199%) | Table 3 omega^2 / Table 9 IIV (%) column |
| `etalka` | 4e-10 (IIV 0.002%) | 4e-10 (IIV 0.002%) | Table 3 / Table 9 IIV (%) column |
| `etalfdepot` | 0.386 | 0.0010278 (IIV 3.206%) | Table 3 omega^2 / Table 9 IIV (%) column |
| `expSd` (sigma, log-additive) | 1.269 | 0.552 | Table 3 / Table 9 |
| Absorption lag time (omitted) | tvTlag 0.000 h | tvTlag 0.000 h | Table 3 / Table 9, Eq. (3) / (8) |

# Virtual cohorts

Each arm of the source studies is simulated with 100 virtual rats. Doses are
per kg, so no body weight is needed. The sampling grid follows the
concentration-time profiles in Figure 4 (0.25-12 h), with a time-zero record
added for NCA.

```{r cohorts}
obs_times <- c(0, 0.25, 0.5, 0.75, 1, 2, 4, 6, 8, 12)
n_per_arm <- 100L

arms_ne <- tribble(
  ~arm,                    ~formulation,     ~admin,  ~dose_mgkg,~FORM_MTX_NANOEMULSION,
  "NP oral 5 mg/kg",       "Nanoparticles",  "oral",  5,      0L,
  "NP IV 5 mg/kg",         "Nanoparticles",  "iv",    5,      0L,
  "NE oral 0.06 mg/kg",    "Nanoemulsions",  "oral",  0.06,   1L,
  "NE IV 0.024 mg/kg",     "Nanoemulsions",  "iv",    0.024,  1L
)

arms_nano <- tribble(
  ~arm,                    ~formulation,     ~admin,  ~dose_mgkg,~FORM_MTX_NANO,
  "Free oral 5 mg/kg",     "Free solution",  "oral",  5,      0L,
  "Free IV 5 mg/kg",       "Free solution",  "iv",    5,      0L,
  "Free oral 0.06 mg/kg",  "Free solution",  "oral",  0.06,   0L,
  "Free IV 0.024 mg/kg",   "Free solution",  "iv",    0.024,  0L,
  "NP oral 5 mg/kg",       "Nanoparticles",  "oral",  5,      1L,
  "NP IV 5 mg/kg",         "Nanoparticles",  "iv",    5,      1L,
  "NE oral 0.06 mg/kg",    "Nanoemulsions",  "oral",  0.06,   1L,
  "NE IV 0.024 mg/kg",     "Nanoemulsions",  "iv",    0.024,  1L
)

# One oral dose into depot or one IV bolus into central per rat, then
# observation records on the central state (Cc is computed from it).
make_events <- function(arms, n, times) {
  subj <- arms |>
    mutate(arm_idx = row_number()) |>
    tidyr::crossing(rep = seq_len(n)) |>
    mutate(id = (arm_idx - 1L) * n + rep)
  doses <- subj |>
    mutate(time = 0, amt = dose_mgkg, evid = 1L,
           cmt = if_else(admin == "oral", "depot", "central"))
  obs <- subj |>
    tidyr::crossing(time = times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
  bind_rows(doses, obs) |>
    select(-arm_idx, -rep) |>
    arrange(id, time, desc(evid)) |>
    as.data.frame()
}

ev_ne <- make_events(arms_ne, n_per_arm, obs_times)
ev_nano <- make_events(arms_nano, n_per_arm, obs_times)
stopifnot(
  length(unique(ev_ne$id)) == nrow(arms_ne) * n_per_arm,
  length(unique(ev_nano$id)) == nrow(arms_nano) * n_per_arm
)
```

# Simulation

```{r simulate}
mod_ne <- rxode2::rxode2(readModelDb("Jeong_2021_methotrexate_nanoemulsion_rat"))
mod_nano <- rxode2::rxode2(readModelDb("Jeong_2021_methotrexate_nanoformulation_rat"))

keep_cols <- c("arm", "formulation", "admin", "dose_mgkg")

# The states are amounts in mg/kg, and the low-dose arms fall to about
# 1e-10 mg/kg by 12 h, below the default absolute tolerance. Tight
# tolerances stop solver noise from producing negative concentrations.
tol <- list(atol = 1e-14, rtol = 1e-10)

rxode2::rxSetSeed(20210709L)
sim_ne <- rxode2::rxSolve(mod_ne, events = ev_ne, keep = keep_cols,
                          atol = tol$atol, rtol = tol$rtol) |>
  as.data.frame()
rxode2::rxSetSeed(20210710L)
sim_nano <- rxode2::rxSolve(mod_nano, events = ev_nano, keep = keep_cols,
                            atol = tol$atol, rtol = tol$rtol) |>
  as.data.frame()
# A few fast-clearing virtual rats still undershoot zero by ~1e-15 ng/mL at
# 8-12 h. Assert that this is noise relative to the peak, then floor it so
# PKNCA does not take logs of negative values.
stopifnot(min(sim_ne$Cc) >= -1e-9 * max(sim_ne$Cc),
          min(sim_nano$Cc) >= -1e-9 * max(sim_nano$Cc))
sim_ne$Cc <- pmax(sim_ne$Cc, 0)
sim_nano$Cc <- pmax(sim_nano$Cc, 0)

# Typical-value (no IIV, no residual) solves on a dense grid for the
# figure replicates and the closed-form check. zeroRe() leaves no omega,
# which rxode2 reports with a warning for a multi-subject event table.
dense_times <- sort(unique(c(seq(0, 12, by = 0.05), obs_times)))
typ_ne <- suppressWarnings(rxode2::rxSolve(
  rxode2::zeroRe(mod_ne),
  events = make_events(arms_ne, 1L, dense_times), keep = keep_cols,
  atol = tol$atol, rtol = tol$rtol
)) |> as.data.frame()
typ_nano <- suppressWarnings(rxode2::rxSolve(
  rxode2::zeroRe(mod_nano),
  events = make_events(arms_nano, 1L, dense_times), keep = keep_cols,
  atol = tol$atol, rtol = tol$rtol
)) |> as.data.frame()
```

## Closed-form check of the encoding

A one-compartment model with first-order absorption has an analytic solution,
so the typical-value solve must reproduce it for every arm. This check shares
its parameters with the solve, so any difference is numerical error and a
tight bound is appropriate. It pins the per-kg units (x 1000 to ng/mL), the
formulation coefficients on V, CL, ka and F, and the dosing compartments.

```{r closed-form}
closed_form <- function(arms, tv, indicator) {
  arms |>
    mutate(
      v = tv$v * (1 + tv$e_v * .data[[indicator]]),
      cl = tv$cl * (1 + tv$e_cl * .data[[indicator]]),
      ka = tv$ka * (1 + tv$e_ka * .data[[indicator]]),
      f = tv$f * (1 + tv$e_f * .data[[indicator]]),
      kel = cl / v
    ) |>
    select(arm, admin, dose_mgkg, v, cl, ka, f, kel)
}

tv_ne <- list(v = 18.832, cl = 9.167, ka = 0.714, f = 0.334,
              e_v = -0.986, e_cl = -0.990, e_ka = 1.552, e_f = 0.193)
tv_nano <- list(v = 14.889, cl = 14.577, ka = 0.582, f = 0.272,
                e_v = 0.429, e_cl = -0.355, e_ka = 10.883, e_f = 4.246)

check_closed_form <- function(typ, params) {
  typ |>
    select(arm, time, Cc) |>
    inner_join(params, by = "arm") |>
    mutate(
      Cc_exact = if_else(
        admin == "iv",
        1000 * dose_mgkg / v * exp(-kel * time),
        1000 * f * dose_mgkg * ka / (v * (ka - kel)) * (exp(-kel * time) - exp(-ka * time))
      ),
      rel_err = abs(Cc - Cc_exact) / max(Cc_exact)
    ) |>
    group_by(arm) |>
    summarise(max_rel_err = max(rel_err), .groups = "drop")
}

cf_ne <- check_closed_form(typ_ne, closed_form(arms_ne, tv_ne, "FORM_MTX_NANOEMULSION"))
cf_nano <- check_closed_form(typ_nano, closed_form(arms_nano, tv_nano, "FORM_MTX_NANO"))
knitr::kable(bind_rows(
  mutate(cf_ne, model = "NP-vs-NE"),
  mutate(cf_nano, model = "Free-vs-nano")
) |> select(model, arm, max_rel_err), digits = 8,
caption = "Largest relative difference between the typical-value solve and the analytic one-compartment solution (scaled by each arm's peak).")
stopifnot(max(cf_ne$max_rel_err) < 1e-4, max(cf_nano$max_rel_err) < 1e-4)
```

# Replicate published figures

## Figure 4 -- dose-normalised profiles, nanoparticles vs nanoemulsions

Figure 4 plots the mean dose-normalised concentration (ng/mL per mg/kg) of the
two nanoformulations after oral (A) and intravenous (B) dosing. The published
means run roughly 6 ng/mL per mg/kg at the nanoparticle oral peak versus about
1400 for the nanoemulsions, and about 140 versus 7500 at 0.25 h after IV
dosing (Table 6 Cmax/Dose).

```{r figure-4, fig.cap = "Replicates Figure 4 of Jeong 2021 with the nanoparticle-vs-nanoemulsion model: typical-value dose-normalised methotrexate concentration after oral (left) and IV (right) dosing."}
typ_ne |>
  filter(time > 0) |>
  mutate(Cc_dn = Cc / dose_mgkg,
         route_label = if_else(admin == "oral", "Oral", "Intravenous")) |>
  ggplot(aes(time, Cc_dn, colour = formulation)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~route_label) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Dose-normalised Cc (ng/mL per mg/kg)",
       colour = NULL)
```

The model reproduces the direction and the roughly 100- to 200-fold separation
between the formulations. It does not reproduce the early IV peak (for example,
the nanoparticle C0 of 1344.57 ng/mL in Table 6 against a model-predicted
265 ng/mL). A one-compartment model cannot represent the rapid initial decline
of an IV profile. The authors state that two- and three-compartment models
"could not be applied to multiple subjects".

## Figure 9 -- visual predictive check by formulation

```{r figure-9, fig.cap = "Replicates Figure 9 of Jeong 2021: simulated 5th, 50th and 95th percentiles of plasma methotrexate for the nanoparticle-vs-nanoemulsion model, pooled over oral and IV dosing within each formulation, with residual error added."}
expSd_ne <- rxode2::rxode(mod_ne)$theta[["expSd"]]
set.seed(20210711L)
vpc_ne <- sim_ne |>
  filter(time > 0) |>
  mutate(obs = Cc * exp(expSd_ne * rnorm(n()))) |>
  group_by(formulation, time) |>
  summarise(q05 = quantile(obs, 0.05), q50 = median(obs), q95 = quantile(obs, 0.95),
            .groups = "drop")

ggplot(vpc_ne, aes(time)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = q50), colour = "steelblue4", linewidth = 0.8) +
  facet_wrap(~formulation) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Plasma methotrexate (ng/mL)")
```

Figure 9 plots concentrations on a natural-log axis. The nanoparticle panel
runs from about e^6.5 ng/mL at 0.25 h (the IV arm) down to about e^0 to e^1.5
at 12 h. The nanoemulsion panel runs from about e^4 to e^5 early down to about
e^0 to e^1 at 12 h. In the simulation the nanoparticle 95th percentile is
e^6.3 at 0.25 h and the 12 h median is e^-0.2. For the nanoemulsions the
0.25 h median is e^3.9 and the 12 h median is e^0.4, the same ranges as the
published panels.

# PKNCA validation

```{r pknca}
run_nca <- function(sim, events) {
  conc <- sim |>
    filter(!is.na(Cc)) |>
    select(id, time, Cc, arm)
  dose_df <- events |>
    filter(evid == 1L) |>
    select(id, time, amt, arm)
  intervals <- data.frame(
    start = 0, end = Inf,
    cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE,
    half.life = TRUE
  )
  PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(conc, Cc ~ time | arm + id),
    PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id),
    intervals = intervals
  ))
}

nca_ne <- run_nca(sim_ne, ev_ne)
nca_nano <- run_nca(sim_nano, ev_nano)
```

## Comparison against published NCA

Table 6 gives the NCA of the nanoformulation arms (mean of 5 rats). The
free-solution arms are shown only as bar charts in Figures 1 and 2; the values
below for those arms were read off the bar heights by the maintainers and are
approximate. The simulated values are medians of 100 virtual rats per arm.
Cmax is compared for the oral arms only. For IV bolus doses the paper's Cmax
is the first sample (0.25 h), which a one-compartment model does not
represent (see Figure 4 above).

```{r nca-reference}
ref_table6 <- tribble(
  ~arm,                  ~PPTESTCD,     ~PPORRES,
  "NP oral 5 mg/kg",     "cmax",        31.19,
  "NP oral 5 mg/kg",     "tmax",        0.92,
  "NP oral 5 mg/kg",     "auclast",     142.05,
  "NP oral 5 mg/kg",     "aucinf.obs",  148.44,
  "NP oral 5 mg/kg",     "half.life",   2.59,
  "NE oral 0.06 mg/kg",  "cmax",        81.72,
  "NE oral 0.06 mg/kg",  "tmax",        1.35,
  "NE oral 0.06 mg/kg",  "auclast",     288.35,
  "NE oral 0.06 mg/kg",  "aucinf.obs",  291.34,
  "NE oral 0.06 mg/kg",  "half.life",   1.58,
  "NP IV 5 mg/kg",       "auclast",     720.15,
  "NP IV 5 mg/kg",       "aucinf.obs",  722.53,
  "NP IV 5 mg/kg",       "half.life",   1.62,
  "NE IV 0.024 mg/kg",   "auclast",     268.94,
  "NE IV 0.024 mg/kg",   "aucinf.obs",  300.56,
  "NE IV 0.024 mg/kg",   "half.life",   6.38
)

# Approximate values read off Figure 1 (free solution, 5 mg/kg) and Figure 2
# (free solution, 0.06 mg/kg oral / 0.024 mg/kg IV) by the maintainers.
ref_fig12 <- tribble(
  ~arm,                    ~PPTESTCD,     ~PPORRES,
  "Free oral 5 mg/kg",     "cmax",        17,
  "Free oral 5 mg/kg",     "aucinf.obs",  62,
  "Free oral 5 mg/kg",     "half.life",   1.55,
  "Free IV 5 mg/kg",       "aucinf.obs",  378,
  "Free IV 5 mg/kg",       "half.life",   1.6,
  "Free oral 0.06 mg/kg",  "cmax",        12,
  "Free oral 0.06 mg/kg",  "aucinf.obs",  29,
  "Free oral 0.06 mg/kg",  "half.life",   1.1,
  "Free IV 0.024 mg/kg",   "aucinf.obs",  94,
  "Free IV 0.024 mg/kg",   "half.life",   0.85
)

nca_units <- c(cmax = "ng/mL", tmax = "h", auclast = "ng*h/mL",
               aucinf.obs = "ng*h/mL", half.life = "h")
```

### Nanoparticle-vs-nanoemulsion model

```{r nca-compare-ne}
cmp_ne <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_ne, reference = ref_table6, by = "arm",
  units = nca_units, tolerance_pct = 20
)
knitr::kable(cmp_ne, caption = "Nanoparticle-vs-nanoemulsion model: simulated median vs Jeong 2021 Table 6. * differs from the reference by more than 20%.")
```

### Free-solution-vs-nanoformulation model

```{r nca-compare-nano}
cmp_nano <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_nano, reference = bind_rows(ref_fig12, ref_table6), by = "arm",
  units = nca_units, tolerance_pct = 20
)
knitr::kable(cmp_nano, caption = "Free-solution-vs-nanoformulation model: simulated median vs Jeong 2021 Figures 1-2 (free solution, approximate) and Table 6 (nanoformulations). * differs from the reference by more than 20%.")
```

```{r nca-assert}
auc_pct <- function(nca, ref, arms) {
  sim_med <- as.data.frame(nca$result) |>
    filter(PPTESTCD == "aucinf.obs", arm %in% arms) |>
    group_by(arm) |>
    summarise(sim = median(PPORRES, na.rm = TRUE), .groups = "drop")
  ref |>
    filter(PPTESTCD == "aucinf.obs", arm %in% arms) |>
    inner_join(sim_med, by = "arm") |>
    mutate(pct = 100 * (sim - PPORRES) / PPORRES)
}

# NP-vs-NE model: every arm's median AUCinf within 40% of Table 6. The
# typical-value predictions sit 11-25% from the published means, so a
# mis-transcribed clearance, bioavailability or unit would break this.
chk_ne <- auc_pct(nca_ne, ref_table6, arms_ne$arm)
chk_ne
stopifnot(nrow(chk_ne) == 4L, all(abs(chk_ne$pct) < 40))

# Free-vs-nano model: only the 5 mg/kg IV arms are within the model's reach
# (see Assumptions); check those.
ref_all <- bind_rows(ref_fig12, ref_table6)
chk_nano <- auc_pct(nca_nano, ref_all, c("Free IV 5 mg/kg", "NP IV 5 mg/kg"))
chk_nano
stopifnot(nrow(chk_nano) == 2L, all(abs(chk_nano$pct) < 40))

# ...and the published model's known failure on the low-dose arms is pinned
# too: their median AUCinf is below a tenth of the observed value (about a
# hundredth at the typical values), so a change that silently "fixed" these
# arms would mean the model no longer matches Table 3.
low_dose <- c("Free oral 0.06 mg/kg", "Free IV 0.024 mg/kg",
              "NE oral 0.06 mg/kg", "NE IV 0.024 mg/kg")
chk_low <- auc_pct(nca_nano, ref_all, low_dose)
chk_low
stopifnot(nrow(chk_low) == 4L, all(chk_low$sim / chk_low$PPORRES < 0.1))
```

The nanoparticle-vs-nanoemulsion model is close to Table 6 on exposure (AUC
within 11-25%). It misses the nanoemulsion IV half-life (about 2 h simulated
against 6.38 h). This follows from the model's own typical values: `CL/V` for
the nanoemulsions is `0.0917 / 0.264 = 0.35 1/h`, which is a 2.0 h half-life.
Table 6 reports a non-compartmental volume of 0.748 L/kg for this arm, about
three times the model's `V`. The oral Tmax is coarse because the sampling grid
jumps from 1 h to 2 h.

The free-solution-vs-nanoformulation model reproduces exposure only for the
two 5 mg/kg IV arms. It over-predicts the 5 mg/kg oral arms, because its
nanoformulation bioavailability is 1.43 and its free-solution bioavailability
(0.272) is above the roughly 16% the authors report in Figure 1F. It
under-predicts the low-dose arms (free solution at 0.06 / 0.024 mg/kg and the
nanoemulsions) by about two orders of magnitude. See the first item under
Assumptions and deviations.

# Assumptions and deviations

- **The free-solution-vs-nanoformulation model has no dose or study effect.**
  Its free-solution arms were dosed at 5 mg/kg in the nanoparticle study and
  at 0.06 mg/kg oral / 0.024 mg/kg IV in the nanoemulsion study. The paper
  gives these doses in the Figure 2 legend, while the Figure 2 caption says
  5 mg/kg. In the source data the two free-solution arms differ roughly
  50-fold in clearance: about 14 L/h/kg in Figure 1C against 0.26 L/h/kg in
  Figure 2C. The model has one typical clearance for all free-solution doses
  (14.577 L/h/kg) and one nanoformulation multiplier that pools nanoparticles
  and nanoemulsions, so it cannot describe both dose levels. The authors note
  that the fit was limited by "the diversity of dosages, and the limited
  covariate values" (Section 3.2). The packaged model reproduces Table 3
  exactly. It is not adjusted to fit the low-dose arms. Only the 5 mg/kg IV
  arms are reproduced (see the NCA comparison), and the
  nanoparticle-vs-nanoemulsion model is the better description of either
  nanoformulation.
- **Absorption lag time omitted.** Both final models estimate `Tlag` (with
  IIV), but Tables 3 and 9 print `tvTlag = 0.000 h` with SE 0.000, so the
  estimate is below 0.0005 h (under 2 seconds). The authors call it "close to
  0" and "not very important to the interpretation" (Section 3.2). A zero lag
  cannot be written on the log scale, and a lag this short has no effect on
  any simulated concentration, so the models omit the lag and its IIV
  (0.052%) rather than invent a nonzero value.
- **Near-zero IIV variances back-solved from the IIV (%) column.** The
  omega^2 column prints three decimals, which reads 0.000 for V and Ka in
  both models and gives only one significant figure for CL and F in the
  nanoparticle-vs-nanoemulsion model. The IIV (%) column is
  `100 * sqrt(omega^2)` (Table 3: `sqrt(0.338) = 58.14%` against the printed
  58.130%), so those variances are taken as `(IIV% / 100)^2`. The two
  well-resolved Table 3 variances (CL 0.338, F 0.386) are used as printed.
- **Residual error is SD on the log scale.** Phoenix NLME reports the
  log-additive sigma as a standard deviation, and the tables label it sigma
  rather than sigma^2, so `expSd` takes the printed value directly.
- **Bioavailability above 1.** In the free-solution-vs-nanoformulation model
  the typical nanoformulation F is `0.272 * (1 + 4.246) = 1.43`. This is the
  published estimate, inflated by the pooled low-dose nanoemulsion arms. It is
  encoded as published and not capped.
- **Per-kg units.** Doses are in mg/kg, V in L/kg and CL in L/h/kg as in the
  paper, and `Cc` is reported in ng/mL. No body weight enters the models.
- **Figure 1/2 reference values are approximate.** The free-solution NCA is
  published only as bar charts. The values in the comparison table were read
  off the bar heights and carry an uncertainty of a few percent.
- **Figure 1 and Table 6 disagree for the nanoparticle IV arm.** Table 6 gives
  a half-life of 1.62 h and V of 16.1 L/kg. Figure 1B/1D show about 4.9 h and
  56 L/kg. The Table 6 values are used as the reference.
- **Virtual cohort size.** 100 rats per arm (the source studies used 5), which
  gives stable medians without exceeding the vignette time budget.
