---
title: "Tiropramide (Jeong 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tiropramide (Jeong 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Jeong_2020_tiropramide"))$reference`
* Article: <https://doi.org/10.3390/pharmaceutics12040374> (open access)

# Population

Jeong et al. (2020) re-analysed the reference-formulation period of a
randomised, single-dose, two-way crossover bioequivalence study of
tiropramide 100 mg tablets in 24 healthy Korean men (Section 2.1). Age
ranged from 19 to 29 years (median 23), body weight from 52.5 to 82.8 kg
(median 66.8), and serum total protein from 6.7 to 8.3 g/dL (median 7.6;
Table 1). Plasma was sampled pre-dose and at 0.33, 0.67, 1, 1.5, 2, 2.5,
3, 4, 6, 8 and 12 h after a fasted oral dose, giving 288 concentrations
(Section 3.1) measured by column-switching semi-micro HPLC with an LLOQ of
2 ng/mL (Section 3.3). ABCB1, CYP2D6, OCT2 and PEPT1 genotypes were
determined for every subject (Table 2) and screened as covariates, but
none was retained.

The final model (Section 3.5, Figure 2, Equation 1) is one-compartment
with first-order elimination. The oral dose enters "depot 1" and, after an
absorption lag time, transfers to "depot 2" at the first absorption rate
constant Ka1, from which it enters the central compartment at Ka2. Serum
total protein enters CL/F and V/F as a linear effect centred at the
7.6 g/dL median. Inter-individual variability is exponential on all five
structural parameters, and the residual error is additive on
log-transformed concentrations. The model was estimated in Phoenix NLME
8.1 with FOCE-ELS.

```{r population}
rxode2::rxode2(readModelDb("Jeong_2020_tiropramide"))$population
```

# Source trace

| Quantity | Value | Source location |
|---|---|---|
| Structure: depot 1 -> depot 2 -> central, lag on depot 1 | -- | Section 3.5 text; Figure 2 schematic and caption |
| Covariate equations (linear, centred at 7.6 g/dL) | -- | Equation 1 |
| Ka1 (`lktr`) | 3.187 1/h | Table 6, final model, tvKa1 |
| Ka2 (`lka`) | 3.183 1/h | Table 6, final model, tvKa2 |
| V/F (`lvc`) | 1,889,250.002 mL = 1889.25 L | Table 6, final model, tvV/F |
| CL/F (`lcl`) | 466,711.101 mL/h = 466.711 L/h | Table 6, final model, tvCL/F |
| Tlag (`ltlag`) | 0.196 h | Table 6, final model, tvTlag |
| Total protein on CL/F (`e_tpro_cl`) | -0.804 per g/dL | Table 6, dCl/FdTotalproteins; sign confirmed by Table 7 |
| Total protein on V/F (`e_tpro_vc`) | -1.049 per g/dL | Table 6, dV/FdTotalproteins; sign confirmed by Table 7 |
| omega^2 Ka1 / Ka2 | 0.557 / 0.556 | Table 6, final model |
| omega^2 V/F / CL/F / Tlag | 0.326 / 0.160 / 0.107 | Table 6, final model |
| sigma (`expSd`) | 0.357 | Table 6, final model; log-additive form in Section 2.6 |
| Dose, sampling times, LLOQ | 100 mg; 0-12 h; 2 ng/mL | Sections 2.1 and 3.3 |

The Table 6 IIV (%) column equals `100 * sqrt(omega^2)` (for example
`sqrt(0.326) = 0.571`, printed as 57.118%), which confirms that the
omega column holds variances. The residual row is printed as `sigma` (not
`sigma^2`) alongside `omega^2` rows and is used as the log-scale standard
deviation.

# Load the model

```{r load-model}
mod <- rxode2::rxode2(readModelDb("Jeong_2020_tiropramide"))
mod_typical <- rxode2::zeroRe(mod)
```

# Typical-value checks

For a first-order-elimination model the area under the curve after a
single oral dose is exactly `Dose / (CL/F)`, whatever the absorption
chain looks like, and the terminal half-life is `log(2) * V / CL` because
both absorption rate constants (about 3.2 1/h) are much faster than
elimination (`CL/V` = 0.247 1/h). At the median total protein of
7.6 g/dL (76 g/L) the expected values are
`100 mg / 466.711 L/h = 214.27 ng*h/mL` and
`log(2) * 1889.25 / 466.711 = 2.806 h`.

```{r typical}
obs_dense <- c(seq(0, 2, by = 0.02), seq(2.05, 72, by = 0.05))

ev_typ <- rxode2::et(amt = 100, cmt = "depot") |>
  rxode2::et(obs_dense, cmt = "central") |>
  as.data.frame()
ev_typ$TPRO <- 76

sim_typ <- as.data.frame(rxode2::rxSolve(mod_typical, events = ev_typ))

auc_typ <- sum(diff(sim_typ$time) *
  (utils::head(sim_typ$Cc, -1) + utils::tail(sim_typ$Cc, -1)) / 2)
auc_expected <- 100 / 466.711101 * 1000

term <- sim_typ[sim_typ$time >= 24, ]
thalf_typ <- log(2) / -stats::coef(stats::lm(log(Cc) ~ time, data = term))[["time"]]
thalf_expected <- log(2) * 1889.250002 / 466.711101

c(
  auc_sim = auc_typ, auc_expected = auc_expected,
  thalf_sim = thalf_typ, thalf_expected = thalf_expected
)

stopifnot(
  # Same drawn parameters on both sides (typical values): pure numerical error.
  abs(auc_typ / auc_expected - 1) < 0.01,
  abs(thalf_typ / thalf_expected - 1) < 0.01
)
```

The lag time delays the first appearance of drug in plasma: the typical
concentration is exactly zero until `Tlag = 0.196 h`.

```{r lag-check}
stopifnot(
  all(sim_typ$Cc[sim_typ$time < 0.19] == 0),
  sim_typ$Cc[sim_typ$time == 0.3] > 0
)
```

## Total-protein effect

Equation 1 scales CL/F and V/F by `1 + (TP - 7.6) * slope`. Across the
observed total-protein range (6.7-8.3 g/dL) this moves CL/F between
1.72-fold and 0.44-fold, and V/F between 1.94-fold and 0.27-fold, of the
median-subject value. Higher total protein therefore means lower CL/F,
lower V/F and higher exposure, consistent with the negative correlations
shown in Figure 3.

```{r tpro-effect, fig.width = 7, fig.height = 4}
tp_grid <- c(6.7, 7.2, 7.6, 8.0, 8.3)
sim_tp <- dplyr::bind_rows(lapply(tp_grid, function(tp) {
  ev <- rxode2::et(amt = 100, cmt = "depot") |>
    rxode2::et(seq(0, 12, by = 0.05), cmt = "central") |>
    as.data.frame()
  ev$TPRO <- tp * 10
  as.data.frame(rxode2::rxSolve(mod_typical, events = ev)) |>
    dplyr::mutate(tp_gdL = tp)
}))

ggplot(sim_tp, aes(time, Cc, colour = factor(tp_gdL))) +
  geom_line() +
  labs(
    x = "Time after dose (h)",
    y = "Tiropramide (ng/mL)",
    colour = "Total protein (g/dL)",
    title = "Typical profiles across the observed total-protein range, 100 mg"
  ) +
  theme_bw()

tp_auc <- sim_tp |>
  dplyr::group_by(tp_gdL) |>
  dplyr::summarise(
    auc_0_12 = sum(diff(time) * (utils::head(Cc, -1) + utils::tail(Cc, -1)) / 2),
    .groups = "drop"
  )
tp_auc |>
  dplyr::rename("Total protein (g/dL)" = tp_gdL, "AUC0-12 (ng*h/mL)" = auc_0_12) |>
  knitr::kable(digits = 1)
```

# Virtual cohort

A cohort of 200 virtual healthy men receives a single 100 mg oral dose.
Total protein is drawn from a normal distribution with the Table 1 mean
and SD (7.59 +/- 0.41 g/dL), truncated to the observed 6.7-8.3 g/dL range
(the linear covariate form becomes non-positive for V/F above 8.55 g/dL,
so extrapolation is not supported). Observations follow the study's
sampling schedule.

```{r cohort}
rxode2::rxSetSeed(20200418)
set.seed(20200418)
n_sub <- 200
tp_draw <- numeric(0)
while (length(tp_draw) < n_sub) {
  x <- rnorm(n_sub, 7.59, 0.41)
  tp_draw <- c(tp_draw, x[x >= 6.7 & x <= 8.3])
}
cohort <- tibble::tibble(id = seq_len(n_sub), TPRO = tp_draw[seq_len(n_sub)] * 10)

samp_times <- c(0, 0.33, 0.67, 1, 1.5, 2, 2.5, 3, 4, 6, 8, 12)
ev_one <- rxode2::et(amt = 100, cmt = "depot") |>
  rxode2::et(sort(unique(c(samp_times, seq(0, 12, by = 0.1)))), cmt = "central") |>
  as.data.frame()

ev_all <- dplyr::bind_rows(lapply(seq_len(n_sub), function(i) {
  e <- ev_one
  e$id <- cohort$id[i]
  e$TPRO <- cohort$TPRO[i]
  e
}))

sim <- as.data.frame(rxode2::rxSolve(mod, events = ev_all, keep = "TPRO"))
```

## Visual predictive check (Figure 5)

```{r vpc, fig.width = 7, fig.height = 4.5}
vpc <- sim |>
  dplyr::filter(time > 0) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    p05 = quantile(sim, 0.05),
    p50 = quantile(sim, 0.50),
    p95 = quantile(sim, 0.95),
    .groups = "drop"
  )

ggplot(vpc, aes(time)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = p50), colour = "steelblue") +
  geom_hline(yintercept = 2, linetype = "dotted") +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)",
    y = "Tiropramide (ng/mL, log scale)",
    title = "Simulated median and 90% prediction interval, 100 mg single dose",
    caption = "Replicates the layout of Figure 5 of Jeong 2020. Dotted line: LLOQ 2 ng/mL."
  ) +
  theme_bw()
```

# PKNCA validation

NCA is run per subject on the individual predictions (`Cc`, without
residual error) at the study's sampling times, with concentrations below
the 2 ng/mL LLOQ removed, as they would have been in the observed data.

```{r pknca}
nca_conc <- sim |>
  dplyr::filter(time %in% samp_times) |>
  dplyr::mutate(Cc = ifelse(time > 0 & Cc < 2, NA_real_, Cc)) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(treatment = "100 mg oral") |>
  dplyr::select(id, time, Cc, treatment)

nca_dose <- cohort |>
  dplyr::transmute(id, time = 0, amt = 100, treatment = "100 mg oral")

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ time | treatment + id,
  concu = "ng/mL", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(nca_dose, amt ~ time | treatment + id,
  doseu = "mg", route = "extravascular"
)
intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE
)
nca_res <- suppressWarnings(PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
))
```

Jeong 2020 reports arithmetic means for the observed NCA (Section 3.4)
and for the same parameters derived from the final population model's
individual predictions (Section 3.5), so the simulated per-subject values
are also summarised by their arithmetic mean before comparison.

```{r nca-compare}
sim_means <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "aucinf.obs", "half.life")) |>
  dplyr::group_by(PPTESTCD) |>
  dplyr::summarise(PPORRES = mean(PPORRES, na.rm = TRUE), .groups = "drop")

sim_long <- dplyr::bind_rows(
  sim_means |> dplyr::mutate(comparison = "Observed NCA (Section 3.4)"),
  sim_means |> dplyr::mutate(comparison = "Final-model predictions (Section 3.5)")
)

published <- tibble::tribble(
  ~comparison,                              ~cmax, ~tmax, ~auclast, ~aucinf.obs, ~half.life,
  "Observed NCA (Section 3.4)",             69.07,  1.74,   254.31,      280.34,       3.41,
  "Final-model predictions (Section 3.5)",  54.89,  1.65,   242.73,      260.92,       2.73
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_long,
  reference = published,
  by = "comparison",
  units = c(cmax = "ng/mL", tmax = "h", auclast = "ng*h/mL", aucinf.obs = "ng*h/mL", half.life = "h")
)
knitr::kable(cmp, caption = "Mean NCA parameters: simulated cohort vs Jeong 2020.")
```

```{r nca-assert}
sim_auc <- sim_means$PPORRES[sim_means$PPTESTCD == "auclast"]
sim_cmax <- sim_means$PPORRES[sim_means$PPTESTCD == "cmax"]
stopifnot(
  # A mis-transcribed CL/F, V/F or unit (mL vs L) moves these by far more.
  abs(sim_auc / 242.73 - 1) < 0.25,
  abs(sim_cmax / 54.89 - 1) < 0.35
)
```

Tmax, AUC and Cmax agree with the paper's final-model-derived values
within about 12%. Two rows exceed the 20% flag:

* **Cmax against the observed NCA.** The observed mean Cmax
  (69.07 +/- 59.74 ng/mL) is higher and far more variable than both the
  paper's own model-derived value (54.89 +/- 46.18 ng/mL) and the
  simulation. It is driven by a few very high individual peaks (Figure 1;
  the two PEPT1 GC subjects averaged 181 ng/mL, Table 3). The simulation
  sits within 10% of the model-derived value, which is the like-for-like
  comparison.
* **Half-life against the model-derived value.** The typical-subject
  half-life is 2.81 h (checked exactly above), close to the paper's
  model-derived mean of 2.73 h. The simulated cohort mean is higher
  (about 3.5 h) because independent log-normal IIV on V/F (omega^2 0.326)
  and CL/F (0.160) makes the mean of `log(2) * V / CL` larger than its
  typical value (a factor of about `exp((0.326 + 0.160) / 2) = 1.28`
  before the total-protein effect). The paper's value was computed from
  the 24 subjects' empirical Bayes estimates, which do not carry the full
  population variance. The simulated mean agrees with the observed NCA
  half-life (3.41 h) within 3%.

The AUClast values are slightly lower than the paper's because the NCA
uses the 12 h sampling schedule with concentrations below the 2 ng/mL
LLOQ removed; AUC0-inf, which extrapolates the tail, is within 5% of the
model-derived value.

# Assumptions and deviations

* **Parameter naming.** The paper's Ka1 moves drug from depot 1 to
  depot 2 and Ka2 from depot 2 to the central compartment (Figure 2
  caption). They are encoded as `lktr` (a single transit step,
  `depot -> transit1`) and `lka` (`transit1 -> central`), respectively.
  The lag time is applied to the dosing compartment (`alag(depot)`),
  matching the caption's "delay time for the drug to move from depot 1 to
  depot 2".
* **Covariate units.** Total protein is supplied in the canonical g/L
  (`TPRO`); the model divides by 10 internally so that Equation 1's g/dL
  slopes and 7.6 g/dL centring apply unchanged.
* **Sign of the total-protein slopes.** The PDF prints both slopes as
  negative in Table 6 and Table 7 (-0.804 on CL/F, -1.049 on V/F),
  consistent with the negative correlations in Figure 3 and the
  Discussion ("higher total protein levels in the blood mean a smaller
  distribution ... and a lower excretion").
* **Validity range.** The linear form in Equation 1 is only valid within
  the observed total-protein range. V/F reaches zero at 8.55 g/dL and
  CL/F at 8.84 g/dL; the virtual cohort is therefore truncated to
  6.7-8.3 g/dL.
* **Residual error.** The Table 6 `sigma` = 0.357 is taken as the SD of
  the additive log-scale error (`Cc ~ lnorm(expSd)`), following the
  table's convention of printing variances as `omega^2` and the residual
  as `sigma`.
* **Unit conversion.** Table 6 reports V/F in mL and CL/F in mL/h; these
  are converted to L and L/h and the concentration is scaled by 1000 to
  give ng/mL from a mg dose.
* **Screened covariates.** BMI, ABCB1 (1236C>T, 2677G>T/A, 3435C>T),
  CYP2D6 (*1/*10), OCT2 808G>T and PEPT1 1287G>C were tested (Table 5)
  but not retained; the tested ones are documented in the model's
  `covariatesDataExcluded` metadata.
* **Errata.** No erratum or correction for doi:10.3390/pharmaceutics12040374
  was found in Europe PMC (no comment/correction links on the record) as of 2026-09-26.
