---
title: "IMC-002 anti-CD47 antibody (Jeong 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IMC-002 anti-CD47 antibody (Jeong 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
ui <- rxode2::rxode(readModelDb("Jeong_2025_imc002"))
```

* Citation: `r ui$reference`
* Description: `r ui$description`
* Article: <https://doi.org/10.1111/cts.70321>
* Supplement (Data S1, NONMEM control stream; Figure S1; Table S1):
  <https://www.ebi.ac.uk/europepmc/webservices/rest/PMC12370379/supplementaryFiles>

IMC-002 is a fully human anti-CD47 IgG4 monoclonal antibody. Jeong 2025 fitted a
semi-mechanistic target-mediated drug disposition (TMDD) model to the phase Ia
dose-escalation data and used it to recommend a phase Ib regimen of 20 mg/kg
every three weeks (Q3W).

# Population

Twelve patients with advanced solid tumours (9 hepatocellular carcinoma, 2 breast
cancer, 1 gallbladder cancer) who had failed standard therapy were enrolled in a
single-centre phase Ia open-label dose-escalation study in the Republic of Korea
(NCT05276310). Three patients were assigned to each of four dose levels (5, 10,
20 and 30 mg/kg), given as a 3 h intravenous infusion every two weeks until
progression or withdrawal, yielding 213 free-IMC-002 concentrations. Mean age was
58.8 years (SD 8.0) and mean weight 64.5 kg (SD 10.2); 8 patients were male and 4
female, and all 12 were Korean (Jeong 2025 Table 1; sampling schedule in Table
S1). Weight, sex and age were screened as covariates by stepwise covariate
modelling (forward p = 0.05, backward p = 0.01) and **none was retained**, so the
packaged model carries no covariate effects.

The same information is available programmatically:

```{r population}
str(ui$population, max.level = 1)
```

# Model structure

Four states are solved. `central` and `peripheral1` carry drug **amounts**
(umol); `target` (CD47) and `complex` (the IMC-002-CD47 complex) carry
**concentrations** (umol/L) referred to the central volume. That asymmetry is the
published parameterisation and is why the volume factors differ between the drug
and receptor equations.

Two distinct binding processes act:

* **CD47, in the central compartment, as full second-order TMDD.** The
  association rate is derived as `Kon = Koff / KD,CD47` from two in-house fixed
  constants; CD47 turns over by zero-order synthesis `Ksyn` and first-order
  degradation `Kdeg`, so its drug-free baseline is `Ksyn / Kdeg`. The complex is
  catabolised, which is the nonlinear elimination route.
* **FcRn, in the peripheral space, at quasi-steady state.** Drug taken up at
  `Kup` equilibrates with a fixed total FcRn pool by the QSS quadratic; the bound
  fraction is recycled back to plasma at `Krec`. Because the FcRn pool is finite,
  this limb saturates at high peripheral concentrations.

```{r structure}
cat(vapply(ui$lstExpr, function(e) paste(deparse(e), collapse = "\n"), character(1)),
    sep = "\n")
```

# Source trace

Per-parameter origin is recorded as an in-file comment beside each `ini()` entry
in `inst/modeldb/specificDrugs/Jeong_2025_imc002.R`. Collected here for review.
"Data S1" is the NONMEM control stream supplied as Supporting Information.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lvc` (Vc) | 4.19 L | Table 2, RSE 4%; Data S1 `$THETA` 1 |
| `lksyn` (Ksyn) | 0.0027 umol/L/h | Table 2, RSE 8%; Data S1 `$THETA` 2 |
| `lkdeg` (Kdeg) | 0.0213 1/h, fixed | Table 2 (starred); sec. 3.2, from CD47 half-life (ref. 14); Data S1 `$THETA` 3 FIX |
| `lkd` (KD,CD47) | 0.046 umol/L, fixed | Table 2 (starred); sec. 3.2, in-house binding; Data S1 `$THETA` 4 FIX |
| `lk2` (Koff) | 90.4 1/h, fixed | Table 2 (starred); sec. 3.2, in-house binding; Data S1 `$THETA` 5 FIX |
| `lcl` (CL) | 0.0115 L/h | Table 2, RSE 12%; Data S1 `$THETA` 6 |
| `lvp` (Vp) | 85.1 L | Table 2, RSE 18%; Data S1 `$THETA` 7 |
| `lkup` (Kup) | 0.0097 1/h | Table 2, RSE 5%; Data S1 `$THETA` 8 |
| `lkss` (KD,FcRn) | 0.117 umol/L, fixed | Table 2 (starred); Data S1 `$THETA` 9 FIX ("KSS1") |
| `lc_fcrn_t` (Rtot,FcRn) | 0.291 umol/L, fixed | Table 2 (starred); sec. 3.2, from Li 2018 (ref. 15); Data S1 `$THETA` 10 FIX |
| `lkrec` (Krec) | 0.0142 1/h | Table 2, RSE 11%; Data S1 `$THETA` 11 |
| `lkint` (Kel,D-R complex) | 0.0188 1/h | Table 2, RSE 13%; Data S1 `$THETA` 12 |
| `expSd` | 0.119 log units | Table 2 row "Proportional error", RSE 9%; Data S1 `$THETA` 13 |
| IIV variances | 0.037 (Vc), 0.319 (CL) | Table 2 "Interindividual variability" rows |
| IIV covariance | 0.0199 | Data S1 `$OMEGA BLOCK(2)` only; not printed in Table 2 |
| `Kon = Koff / KD,CD47` | n/a | Data S1 `$PK`; paper sec. 3.2 symbol list |
| `target(0) = Ksyn / Kdeg` | n/a | Data S1 `$PK` (`BASE`); paper sec. 3.2 initial conditions |
| QSS quadratic for free peripheral drug | n/a | Data S1 `$DES` (`DAA` / `Dfree` / `Dcpx`); paper sec. 3.2 equation for `C` |
| `d/dt(central)`, `d/dt(target)`, `d/dt(complex)`, `d/dt(peripheral1)` | n/a | Data S1 `$DES`; paper sec. 3.2 differential equations |

# Unit conversion

The model is parameterised entirely in molar units because the fixed binding
constants (`KD,CD47`, `KD,FcRn`, `Rtot,FcRn`) and the estimated `Ksyn` are all
reported in umol/L. The paper prescribes doses in mg/kg and reports exposures in
ug/mL, so a molecular weight is needed to move between the two. Jeong 2025 does
not report the molecular weight of IMC-002; a representative human IgG4 value is
assumed here and flagged in "Assumptions and deviations".

```{r units}
MW  <- 146000      # g/mol, assumed human IgG4 molecular weight (NOT from the paper)
WT  <- 64.5        # kg, Table 1 mean body weight
TAU <- 21 * 24     # h, Q3W dosing interval
NCYC <- 8L         # cycles simulated; the paper reports steady state by cycle 5
MEC <- 24          # ug/mL, minimum effective concentration (sec. 2.3)

mgkg_to_umol <- function(mgkg) mgkg * WT * 1000 / MW   # mg/kg -> umol for a 64.5 kg patient
umolL_to_ugmL <- function(x) x * MW / 1000             # umol/L -> ug/mL
```

# Structural checks

These are deterministic typical-value checks with no simulated variability, so
they are exact rather than cohort-dependent and can be asserted tightly.

```{r structural-checks}
# Guard: the model must solve as an ODE system. A cl/vc pair can make rxode2
# auto-substitute a linear closed-form solution and silently discard the ODEs.
stopifnot(!isTRUE(ui$linCmt),
          identical(ui$state, c("central", "target", "complex", "peripheral1")))

mod_typ <- rxode2::zeroRe(readModelDb("Jeong_2025_imc002"))
ev_typ <- rxode2::et(amt = mgkg_to_umol(20), dur = 3, ii = TAU,
                     until = TAU * (NCYC - 1), cmt = "central") |>
  rxode2::et(seq(0, TAU * NCYC, by = 0.25), cmt = "central")
sim_typ <- rxode2::rxSolve(mod_typ, ev_typ, returnType = "data.frame",
                           atol = 1e-12, rtol = 1e-10, maxsteps = 1e6)

# 1. Drug-free CD47 baseline must equal Ksyn / Kdeg exactly (Data S1 $PK BASE).
base_expected <- 0.0027 / 0.0213
stopifnot(abs(sim_typ$target[1] / base_expected - 1) < 1e-8)

# 2. Steady-state mass balance. Over one dosing interval at steady state the dose
#    must equal what leaves the system: linear clearance from central plus
#    catabolism of the drug-CD47 complex. This is the cheapest gate that pins the
#    whole ODE system, including the two fixed binding constants.
trap <- function(x, y) sum(diff(x) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)
ss <- sim_typ |>
  dplyr::filter(time >= TAU * (NCYC - 1), time <= TAU * NCYC) |>
  dplyr::arrange(time)
elim_linear  <- trap(ss$time, ss$kel * ss$central)
elim_complex <- trap(ss$time, ss$kint * ss$complex)
elim_total   <- elim_linear + elim_complex
stopifnot(abs(elim_total / mgkg_to_umol(20) - 1) < 0.01)

tibble::tibble(
  Check = c("CD47 baseline = Ksyn/Kdeg",
            "Steady-state mass balance (dose vs eliminated)",
            "Share of elimination via linear CL",
            "Share of elimination via CD47-complex catabolism"),
  Value = c(sprintf("%.6f umol/L", sim_typ$target[1]),
            sprintf("%.4f%% error", 100 * abs(elim_total / mgkg_to_umol(20) - 1)),
            sprintf("%.1f%%", 100 * elim_linear / elim_total),
            sprintf("%.1f%%", 100 * elim_complex / elim_total))
) |>
  knitr::kable(caption = "Deterministic structural checks at 20 mg/kg Q3W steady state.")
```

At 20 mg/kg roughly half the dose is cleared by catabolism of the drug-CD47
complex, consistent with the paper's statement that IMC-002 "is mainly eliminated
through drug-complex catabolism, not in the blood compartment" (sec. 4).

# Virtual cohort

The published simulations drew 1000 subjects per regimen. Rather than resample
randomly, the cohort below is a **deterministic product-quadrature lattice** over
the bivariate log-normal IIV on Vc and CL: 14 stratified quantile points per eta,
transformed by the Cholesky factor of the published OMEGA. This gives 196
subjects per arm (inside the 200-per-arm cap), reproduces the published variances
exactly, and -- unlike random sampling -- is identical on every machine
regardless of the number of solver threads. That removes the usual source of
locally-passing / CI-failing assertions.

```{r cohort}
K <- 14L
z <- stats::qnorm((seq_len(K) - 0.5) / K)
z <- z / sqrt(mean(z^2))   # rescale so the lattice variance is exactly 1
etas <- as.matrix(expand.grid(z_vc = z, z_cl = z)) %*% t(t(chol(ui$omega)))
colnames(etas) <- c("etalvc", "etalcl")

# The lattice reproduces the published OMEGA diagonal exactly.
lattice_var <- diag(stats::var(etas) * (nrow(etas) - 1) / nrow(etas))
stopifnot(max(abs(lattice_var / diag(ui$omega) - 1)) < 1e-8)

doses <- c(5, 10, 20, 30)
arm_levels <- paste0(doses, " mg/kg")

subjects <- lapply(seq_along(doses), function(i) {
  tibble::tibble(
    id     = (i - 1L) * nrow(etas) + seq_len(nrow(etas)),  # disjoint id ranges per arm
    etalvc = etas[, "etalvc"],
    etalcl = etas[, "etalcl"],
    arm    = factor(arm_levels[i], levels = arm_levels),
    amt    = mgkg_to_umol(doses[i])
  )
}) |>
  dplyr::bind_rows()

obs_grid <- unique(c(seq(0, TAU * (NCYC - 1), by = 12),
                     seq(TAU * (NCYC - 1), TAU * NCYC, by = 3)))

events <- dplyr::bind_rows(
  tidyr::crossing(subjects, time = seq(0, TAU * (NCYC - 1), by = TAU)) |>
    dplyr::mutate(evid = 1L, dur = 3, cmt = "central"),
  tidyr::crossing(dplyr::select(subjects, -amt), time = obs_grid) |>
    dplyr::mutate(evid = 0L, amt = NA_real_, dur = NA_real_, cmt = "central")
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

`etalvc` / `etalcl` are supplied per subject in the event table and `omega = NA`
suppresses resampling, so the lattice is used exactly as constructed.

```{r simulate}
sim <- rxode2::rxSolve(
  readModelDb("Jeong_2025_imc002"), events, omega = NA,
  keep = "arm", returnType = "data.frame",
  atol = 1e-10, rtol = 1e-8, maxsteps = 1e5
)
sim <- sim |>
  dplyr::mutate(arm = factor(as.character(arm), levels = arm_levels),
                conc = umolL_to_ugmL(Cc))
stopifnot(!anyNA(sim$conc), all(sim$conc >= 0))
```

# Replicate published figures

```{r figure-3, fig.cap = "Replicates Figure 3 of Jeong 2025: simulated IMC-002 concentration over the steady-state Q3W dosing interval by dose level, with the 24 ug/mL MEC."}
sim |>
  dplyr::filter(time >= TAU * (NCYC - 1), time <= TAU * NCYC) |>
  dplyr::mutate(t_in_tau = time - TAU * (NCYC - 1)) |>
  dplyr::group_by(arm, t_in_tau) |>
  dplyr::summarise(Q05 = stats::quantile(conc, 0.05),
                   Q50 = stats::median(conc),
                   Q95 = stats::quantile(conc, 0.95), .groups = "drop") |>
  ggplot(aes(t_in_tau, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  geom_hline(yintercept = MEC, linetype = "dashed", colour = "grey40") +
  facet_wrap(~arm) +
  scale_y_log10() +
  labs(x = "Time within the Q3W interval (h)", y = "IMC-002 (ug/mL)",
       title = "Steady-state Q3W profiles by dose level",
       caption = "Replicates Figure 3 of Jeong 2025. Dashed line: MEC 24 ug/mL.")
```

Figure 2 of Jeong 2025 is a visual predictive check overlaying the observed phase
Ia concentrations, which are not publicly available (Data Availability
Statement), so only the simulation-only Figure 3 is reproduced here.

The TMDD mechanism itself is not plotted in the paper, but the packaged model
exposes it directly: free CD47 is driven far below its `Ksyn / Kdeg` baseline
while drug is present and recovers between doses.

```{r target-occupancy, fig.cap = "Free CD47 and the IMC-002-CD47 complex over the steady-state interval (typical-value subject, 20 mg/kg Q3W). Not a published figure."}
sim_typ |>
  dplyr::filter(time >= TAU * (NCYC - 1), time <= TAU * NCYC) |>
  dplyr::transmute(t_in_tau = time - TAU * (NCYC - 1),
                   `Free CD47` = target, `IMC-002-CD47 complex` = complex) |>
  tidyr::pivot_longer(-t_in_tau, names_to = "species", values_to = "conc") |>
  ggplot(aes(t_in_tau, conc, colour = species)) +
  geom_line() +
  geom_hline(yintercept = base_expected, linetype = "dashed", colour = "grey40") +
  labs(x = "Time within the Q3W interval (h)", y = "Concentration (umol/L)",
       colour = NULL,
       caption = "Dashed line: drug-free CD47 baseline Ksyn/Kdeg.") +
  theme(legend.position = "bottom")
```

# PKNCA validation

NCA is taken over the final simulated dosing interval, which is at steady state
(the paper reports steady state by cycle 5; cycle 8 is simulated). `auclast` over
that interval is AUCtau and `cmin` is the steady-state trough: the profile
declines monotonically after each infusion, so the minimum sits at one of the two
interval endpoints, and at steady state both are troughs. (The MEC section below
uses the end-of-interval trough explicitly; the two agree to within 2%.)

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::transmute(id, time, Cc = conc, arm)

# Guarantee a time-zero record per subject (Cc = 0 pre-dose) so PKNCA has an anchor.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, arm, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

dose_df <- events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, arm)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)

intervals <- data.frame(
  start   = TAU * (NCYC - 1),
  end     = TAU * NCYC,
  auclast = TRUE,
  cmax    = TRUE,
  cmin    = TRUE,
  tmax    = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_long <- as.data.frame(nca_res)

# A gate that cannot go red is worse than none: confirm every arm/parameter cell
# actually produced a value.
stopifnot(nrow(nca_long) > 0, !anyNA(nca_long$PPORRES))
stopifnot(setequal(unique(nca_long$PPTESTCD), c("auclast", "cmax", "cmin", "tmax")))
```

## Comparison against published NCA

Jeong 2025 Table 3 reports the **mean** of each parameter over its 1000 simulated
subjects. `ncaComparisonTable()` aggregates per-subject results with the median
by default, so the simulated side is pre-aggregated to arm means here to match
the published statistic.

```{r nca_compare}
sim_wide <- nca_long |>
  dplyr::filter(PPTESTCD %in% c("auclast", "cmin")) |>
  dplyr::mutate(arm = factor(as.character(arm), levels = arm_levels)) |>
  dplyr::group_by(arm, PPTESTCD) |>
  dplyr::summarise(value = mean(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = value)

published <- tibble::tibble(
  arm     = factor(arm_levels, levels = arm_levels),
  auclast = c(8689, 23760, 66036, 116776),   # Table 3, AUCtau (ug*h/mL)
  cmin    = c(5.2, 15.6, 58.8, 122.4)        # Table 3, Ctrough (ug/mL)
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = as.data.frame(sim_wide),
  reference = as.data.frame(published),
  by        = "arm",
  units     = c(auclast = "ug*h/mL", cmin = "ug/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp, align = c("l", "l", "r", "r", "r"),
             caption = "Simulated vs. Jeong 2025 Table 3 (Q3W steady state). * differs by >20%.")
attr(cmp, "footnote")
```

```{r nca_gate}
# ncaComparisonTable's "% diff" column is formatted character, so the gate is
# computed numerically from the underlying values.
gate <- published |>
  dplyr::rename(ref_auc = auclast, ref_cmin = cmin) |>
  dplyr::left_join(sim_wide, by = "arm") |>
  dplyr::mutate(pct_auc  = 100 * (auclast / ref_auc - 1),
                pct_cmin = 100 * (cmin / ref_cmin - 1))

# The cohort is a deterministic lattice, so these numbers do not move between
# machines or thread counts. Realised: AUCtau -7.2 / -8.8 / -1.8 / +2.1 %,
# Ctrough -1.9 / -7.1 / +3.0 / +7.8 %. The 20% bound leaves headroom for the
# assumed molecular weight (the one non-paper input) while still going red on a
# mis-transcribed volume, dose or unit -- encoding the drug in mass rather than
# molar units, for instance, moves these by 30 to 95%.
stopifnot(max(abs(gate$pct_auc)) < 20, max(abs(gate$pct_cmin)) < 20)

gate |>
  dplyr::transmute(
    `Dose level`             = arm,
    `AUCtau % diff`          = round(pct_auc, 1),
    `Ctrough % diff`         = round(pct_cmin, 1)
  ) |>
  knitr::kable(caption = "Numeric deviation from Jeong 2025 Table 3.")
```

## Proportion of patients above the MEC

Table 3 also reports the percentage of simulated patients whose steady-state
trough exceeds the 24 ug/mL MEC. This is the criterion the 20 mg/kg Q3W
recommendation rests on.

```{r mec}
mec_sim <- sim |>
  dplyr::filter(time >= TAU * (NCYC - 1), time <= TAU * NCYC) |>
  dplyr::group_by(arm, id) |>
  dplyr::summarise(ctrough = conc[which.max(time)], .groups = "drop") |>
  dplyr::group_by(arm) |>
  dplyr::summarise(simulated = 100 * mean(ctrough > MEC), .groups = "drop")

mec_cmp <- mec_sim |>
  dplyr::mutate(published = c(NA, 15.6, 74.6, 90.5),   # Table 3 "C trough over MEC (%)"
                difference = simulated - published)

# Realised absolute differences: 6.4 / 3.5 / 2.4 percentage points (deterministic).
stopifnot(max(abs(mec_cmp$difference), na.rm = TRUE) < 15)

mec_cmp |>
  dplyr::transmute(`Dose level` = arm,
                   `Simulated % above MEC` = round(simulated, 1),
                   `Published % above MEC` = published,
                   `Difference (points)`   = round(difference, 1)) |>
  knitr::kable(caption = "Percentage of patients with steady-state Q3W trough above 24 ug/mL, vs Jeong 2025 Table 3.")
```

The model reproduces the paper's decision: at 20 mg/kg Q3W roughly three quarters
of patients hold a trough above the MEC, against roughly one in ten at 10 mg/kg.
Table 3 has no entry for 5 mg/kg, and the model puts that arm at 0%.

# Assumptions and deviations

* **Molecular weight (the only non-paper input).** The model is molar throughout,
  but doses are prescribed in mg/kg and exposures reported in ug/mL. Jeong 2025
  never states the molecular weight of IMC-002, so `MW = 146000 g/mol` -- a
  representative value for a human IgG4 monoclonal antibody -- is assumed here to
  convert. Every ug/mL figure in this vignette therefore carries that assumption;
  the packaged model file itself does not, and is dosed in umol.
* **The printed equation for the free peripheral concentration has a typo; Data
  S1 settles it.** Section 3.2 prints
  `C = 0.5 * [(Ctot,p - Ap - KD,FcRn) + sqrt((Ctot,p - Ap - KD,FcRn)^2 + 4*KD,FcRn*Ctot,p)]`,
  in which `Ap` is defined two lines later as the *amount* of IMC-002 in the
  peripheral compartment -- subtracting an amount from a concentration. Data S1
  `$DES` writes the same line as `DAA = A(4)/VP - FcRn - KSS1`, i.e. the middle
  term is the total FcRn concentration `Rtot,FcRn`, which is also what the
  standard quasi-steady-state root requires. The control-stream form is encoded.
* **Complex elimination is reproduced with its published volume factor.** Both
  the printed equation and Data S1 write the complex catabolism term as
  `Kel,D-R complex * ADR47 / Vc`, which is dimensionally inconsistent with the
  rest of that equation (the other terms are umol/L/h; this one is umol/L/h/L).
  The estimate 0.0188 1/h was obtained under that form, so it is reproduced
  verbatim: rescaling it to remove the stray volume would change the model rather
  than correct it. The effective first-order rate on the complex is therefore
  0.0188 / 4.19 = 0.00449 1/h.
* **`Kup` is a rate constant, not a clearance.** Table 2 tags `Kup` as "(L/h)",
  but the equations use it as `Kup * AD` on an amount, which requires 1/h. The
  label in the model file records 1/h.
* **Residual error is log-additive, not proportional.** Table 2 labels the row
  "Proportional error", but Data S1 `$ERROR` fits `IPRED = LOG(A(1)/VC)` with
  `Y = IPRED + W*EPS(1)`, `W = THETA(13)` and `$SIGMA 1 FIX`. That is a
  log-normal residual, encoded as `Cc ~ lnorm(expSd)`. At 0.119 the two forms are
  numerically close. The Table 2 row "Residual variability 1*" is the fixed
  `$SIGMA` scaffolding, not a separate parameter.
* **The IIV covariance comes from the supplement only.** Table 2 prints the two
  variances but no correlation; Data S1 `$OMEGA BLOCK(2)` supplies the
  off-diagonal 0.0199 (correlation 0.183). Its diagonals reproduce Table 2
  exactly, which is what supports treating the block as final rather than initial
  estimates.
* **Where Table 2 and Data S1 disagree slightly, Table 2 wins.** Data S1 carries
  `Ksyn 0.00273` vs Table 2 `0.0027`, `Vp 85` vs `85.1`, and `Kup 0.00968` vs
  `0.0097`; the published final table is used for all three, with the control
  stream supplying structure.
* **Cohort construction.** The published simulations used 1000 randomly sampled
  subjects per regimen; this vignette uses a 196-point deterministic quadrature
  lattice per arm (14 x 14 stratified quantiles of the two etas, rescaled to
  reproduce the published variances exactly). Means over a lattice and over 1000
  random draws are not identical, which contributes part of the residual
  deviation in the comparison tables.
* **Body weight is held at the Table 1 mean (64.5 kg).** Weight was screened and
  not retained as a covariate, so it enters only through the mg/kg dose
  conversion; using the mean weight matches the paper's per-dose-level reporting.
* **No observed data.** The phase Ia concentrations are not publicly available,
  so the visual predictive check of Figure 2 and the goodness-of-fit plots of
  Figure S1 cannot be reproduced. Validation here is against the published
  simulation summary (Table 3) and the model's own internal mass balance.
