---
title: "Cyanidin-3-Glucoside (Jeon 2012)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cyanidin-3-Glucoside (Jeon 2012)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
suppressPackageStartupMessages({
  library(nlmixr2lib)
  library(PKNCA)
  library(rxode2)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Jeon_2012_C3G"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Jeon_2012_C3G"))$description`
* Article (open access): <https://doi.org/10.4196/kjpp.2012.16.4.249>

Jeon 2012 is a brief open-access pharmacokinetic study of cyanidin-3-glucoside
(C3G), the major anthocyanin in the seed coat of the Korean black bean cultivar
*Phaseolus vulgaris* Cheongjakong-3-ho. Twelve healthy adult Korean volunteers
received 1 g of black bean seed coat extract once daily for 14 days, with
intensive PK sampling on Day 1 and Day 14. The structural model fit is a
one-compartment first-order absorption model with an absorption lag time,
log-normal IIV on CL/F and V/F (correlated, rho = 0.883) and on Ka, and a
proportional residual error. No covariate (age, height, weight, sex, creatinine
clearance) cleared the forward-addition / backward-elimination significance
threshold in the final model.

# Population

Jeon 2012 enrolled and dosed 12 healthy adult Korean volunteers (Jeon 2012
Table 1):

* **N = 12** subjects, all completed the 14-day dosing and PK schedule.
* Sex: 6 male / 6 female (50% female).
* Age: median 29.5 years (range 24-44).
* Height: median 166 cm (range 158-185).
* Weight: median 59.6 kg (range 47.0-78.4).
* All Korean; one centre (Clinical Trial Center, Seoul St Mary's Hospital).
* Diet: low-phenolic during the outpatient period to minimise dietary
  anthocyanin background; at least 10 h overnight fast on PK days
  (Day 1, Day 14) and 4 h fasting after dosing.
* Dosing: 1 g of black bean seed coat extract (200 mg x 5 capsules, GMP-
  produced from 20 kg of Cheongjakong-3-ho beans extracted in 80% ethanol)
  once daily at approximately 9 AM for 14 days.

The same information is available programmatically:
`readModelDb("Jeon_2012_C3G")$population`.

# Source trace

The per-parameter origin is recorded as in-file comments next to each `ini()`
entry of `inst/modeldb/specificDrugs/Jeon_2012_C3G.R`. The table below collates
them for review.

| Parameter (nlmixr2lib) | Value | Source location |
|------------------------|-------|-----------------|
| `lka`             | log(9.94)             | Table 3 Ka = 9.94 1/h (%RSE 30.7) |
| `lcl`             | log(3420)             | Table 3 CL/F = 3420 L/h (%RSE 8.68) |
| `lvc`             | log(7280)             | Table 3 V/F = 7280 L (%RSE 6.84) |
| `lalag`           | log(0.217)            | Table 3 ALAG = 0.217 h (%RSE 3.67) |
| `etalcl`+`etalvc` | block(0.080216, 0.052679, 0.044370) | Table 3 omega_CL = 28.9% CV, omega_V = 21.3% CV, rho_CL-V = 0.883 |
| `etalka`          | 0.718145              | Table 3 omega_Ka = 102.5% CV; log(1 + 1.025^2) |
| `propSd`          | 0.213                 | Table 3 prop residual = 21.3% (%RSE 9.30) |
| `d/dt(depot)`     | n/a                   | One-compartment first-order absorption (Jeon 2012 Methods "Mixed effect PK model development" + Results "Final model") |
| `d/dt(central)`   | n/a                   | First-order elimination from central |
| `alag(depot) <- alag1` | n/a              | Absorption lag time ALAG (Jeon 2012 Table 3) |
| `Cc = 1000 * central / vc` | n/a          | Plasma C3G in ng/mL (dose mg / V L * 1000) |

# Virtual cohort

Original observed plasma C3G concentrations are not redistributed with the
package. The cohort below mirrors the Jeon 2012 study design: 12 subjects
receive a single 14.57 mg oral dose of C3G equivalent on Day 1 (the C3G
content of 1 g of black bean seed coat extract, back-calculated from the
published CL/F and observed AUClast as described in the model file). A
second cohort runs the same 14-day daily dosing regimen so a Day-1 vs Day-14
NCA comparison can be drawn against Table 2 of the paper.

```{r cohort}
set.seed(20120401L)

n_per_arm <- 100L

# Per-administration C3G equivalent dose for 1 g of Cheongjakong-3-ho
# black bean seed coat extract. Back-calculated: AUClast_obs x CL/F =
# 0.00426 mg*h/L x 3420 L/h = 14.57 mg.
dose_c3g_mg <- 14.57

# Sampling grid matches the paper: pre-dose and 0.25, 0.5, 1, 1.5, 2, 4,
# 6 h after dosing on the PK days.
sample_grid_h <- c(0.05, 0.25, 0.5, 1, 1.5, 2, 4, 6)

make_cohort_single <- function(id_offset) {
  ids <- id_offset + seq_len(n_per_arm)

  dose_rows <- tibble::tibble(
    id   = ids,
    time = 0,
    amt  = dose_c3g_mg,
    evid = 1L,
    cmt  = "depot"
  )

  obs_rows <- tidyr::expand_grid(id = ids, time = sample_grid_h) |>
    dplyr::mutate(amt = 0, evid = 0L, cmt = "Cc")

  dplyr::bind_rows(dose_rows, obs_rows) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

make_cohort_multi <- function(id_offset) {
  ids <- id_offset + seq_len(n_per_arm)

  dose_days <- seq(0, 24 * 13, by = 24)
  dose_rows <- tidyr::expand_grid(id = ids, time = dose_days) |>
    dplyr::mutate(amt = dose_c3g_mg, evid = 1L, cmt = "depot")

  # Day 1 PK sampling (0-6 h) plus Day 14 PK sampling (312-318 h after
  # the first dose), matching the paper's blood sampling protocol.
  obs_day1_h  <- sample_grid_h
  obs_day14_h <- 24 * 13 + sample_grid_h
  obs_times   <- c(obs_day1_h, obs_day14_h)

  obs_rows <- tidyr::expand_grid(id = ids, time = obs_times) |>
    dplyr::mutate(amt = 0, evid = 0L, cmt = "Cc")

  dplyr::bind_rows(dose_rows, obs_rows) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  make_cohort_single(id_offset = 0L)         |> dplyr::mutate(treatment = "Day 1 (single dose)"),
  make_cohort_multi(id_offset = n_per_arm)   |> dplyr::mutate(treatment = "Day 14 (multi-dose)")
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid", "cmt")])))
```

# Simulation

```{r simulate}
mod <- rxode2::rxode2(readModelDb("Jeon_2012_C3G"))

sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep   = c("treatment")
) |>
  as.data.frame()
```

For the typical-value reference trajectory (deterministic, no IIV) used to
overlay published NCA point estimates, zero out the random effects:

```{r simulate-typical}
mod_typical <- mod |> rxode2::zeroRe()

sim_typical <- rxode2::rxSolve(
  mod_typical,
  events = events,
  keep   = c("treatment")
) |>
  as.data.frame()
```

# Replicate Figure 2 - VPC by study day

Jeon 2012 Figure 2 shows the visual predictive check of the final model
against observed plasma C3G concentrations on Day 1 and Day 14, with 5th /
50th / 95th percentile bands overlaid on observed data. The stochastic
trajectory below reproduces the same VPC structure (band widths and shapes
are determined entirely by the IIV variances reported in Table 3 and the
proportional residual error).

```{r figure-2-vpc, fig.height = 5}
vpc_day <- sim |>
  dplyr::filter(!is.na(Cc), Cc > 0) |>
  dplyr::mutate(
    panel = treatment,
    tad   = ifelse(treatment == "Day 14 (multi-dose)", time - 24 * 13, time)
  ) |>
  dplyr::filter(tad >= 0, tad <= 6) |>
  dplyr::group_by(panel, tad) |>
  dplyr::summarise(
    Q05 = stats::quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = stats::quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = stats::quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(vpc_day, aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(linewidth = 0.7) +
  facet_wrap(~ panel) +
  labs(
    x = "Time after most-recent dose (h)",
    y = "Plasma C3G (ng/mL)",
    title = "Plasma C3G VPC by study day (replicates Jeon 2012 Figure 2)",
    caption = paste("Median (line) and 5th-95th percentile band (shaded);",
                    "n =", n_per_arm, "per arm; C3G equivalent dose",
                    dose_c3g_mg, "mg per administration.")
  )
```

# Replicate typical-value Cmax / Tmax

The deterministic (zero-RE) trajectory locates the typical-value Cmax and
Tmax, which can be compared against the observed Day 1 / Day 14 NCA values in
Jeon 2012 Table 2.

```{r typical-cmax-tmax}
typical_summary <- sim_typical |>
  dplyr::filter(!is.na(Cc), Cc > 0) |>
  dplyr::mutate(
    tad   = ifelse(treatment == "Day 14 (multi-dose)", time - 24 * 13, time)
  ) |>
  dplyr::filter(tad >= 0, tad <= 6) |>
  dplyr::group_by(treatment, id) |>
  dplyr::slice_max(Cc, n = 1, with_ties = FALSE) |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    Cmax_typical_ngmL = round(stats::median(Cc), 3),
    Tmax_typical_h    = round(stats::median(tad), 3),
    .groups = "drop"
  )

knitr::kable(
  typical_summary,
  caption = "Typical-value (zero-RE) Cmax and Tmax by study day."
)
```

# PKNCA validation

Jeon 2012 Table 2 reports noncompartmental Cmax, Tmax, AUClast, and terminal
half-life on Day 1 and Day 14 (mean +/- SD). The PKNCA block below recomputes
the same parameters from the stochastic simulation for side-by-side
comparison.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc), Cc > 0) |>
  dplyr::mutate(
    tad = ifelse(treatment == "Day 14 (multi-dose)", time - 24 * 13, time)
  ) |>
  dplyr::filter(tad >= 0, tad <= 6) |>
  dplyr::transmute(id, time = tad, Cc, treatment)

dose_df <- events |>
  dplyr::filter(evid == 1L) |>
  dplyr::group_by(id, treatment) |>
  dplyr::summarise(time = 0, amt = dose_c3g_mg, .groups = "drop")

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             route = "extravascular")

intervals <- data.frame(
  start     = 0,
  end       = 6,
  cmax      = TRUE,
  tmax      = TRUE,
  auclast   = TRUE,
  half.life = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- PKNCA::pk.nca(nca_data)

nca_summary <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "half.life")) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(
    mean = round(mean(PPORRES, na.rm = TRUE), 3),
    sd   = round(stats::sd(PPORRES, na.rm = TRUE), 3),
    .groups = "drop"
  )

knitr::kable(
  nca_summary,
  caption = paste("Simulated NCA by study day (n =", n_per_arm, "per arm).",
                  "Cmax ng/mL; Tmax h; AUClast ng*h/mL; half-life h.")
)
```

## Comparison against published Table 2 (noncompartmental)

| Quantity (mean +/- SD)   | Day 1 observed (Table 2) | Day 14 observed (Table 2) | Day 1 simulated | Day 14 simulated |
|--------------------------|--------------------------|---------------------------|-----------------|-------------------|
| Cmax (ng/mL)             | 2.10 +/- 0.60            | 2.17 +/- 0.68             | see PKNCA chunk | see PKNCA chunk  |
| Tmax (h)                 | 0.73 +/- 0.29            | 0.75 +/- 0.34             | see PKNCA chunk | see PKNCA chunk  |
| AUClast (ng*h/mL)        | 4.26 +/- 1.40            | 4.85 +/- 1.65             | see PKNCA chunk | see PKNCA chunk  |
| Terminal half-life (h)   | 1.58 +/- 0.33            | 1.47 +/- 0.29             | see PKNCA chunk | see PKNCA chunk  |

The simulated Day 1 and Day 14 PK characteristics should agree with the
published noncompartmental analysis within the expected sampling variability
for the dense-sampling design used in this 12-subject study. The model
predicts no clinically meaningful accumulation between Day 1 and Day 14
(consistent with Jeon 2012 Results: paired t-test p > 0.05 for all NCA
comparisons between days), driven by the very short terminal half-life
(~1.5 h) relative to the 24 h dosing interval.

# Assumptions and deviations

- **C3G equivalent dose (14.57 mg per 1 g of extract) is back-calculated, not
  reported.** The paper administers 1 g of black bean seed coat extract daily
  but does not report the C3G content per gram of extract. The published
  CL/F and observed Day-1 AUClast jointly imply an effective C3G dose of
  approximately `AUClast x CL/F = 0.00426 mg*h/L x 3420 L/h = 14.57 mg`,
  consistent with the published model fit under the standard NONMEM
  convention that F = 1 is implicit when no separate bioavailability
  parameter is estimated. This value is treated as a dose-unit
  documentation choice: the packaged model takes dose in mg of C3G
  equivalent, and 14.57 mg per administration reproduces the observed Day 1
  and Day 14 NCA values. Anthocyanin oral bioavailability is reported in
  the literature as approximately 1-2% of the ingested mass, so a C3G
  equivalent dose of approximately 1.5% of the 1 g extract mass is
  plausible.

- **IOV vs BSV labelling for omega_Ka.** Jeon 2012 Methods describes both
  between-subject variability (eta on log-CL / log-V / log-Ka / log-ALAG)
  and inter-occasion variability (kappa on log-CL / log-V / log-Ka), and
  the final-model prose states that "inter-occasional variability between
  Day 1 and Day 14 was tested for all PK parameters, but was finally found
  relevant only for absorption rate constant (Ka)." Table 3 reports a
  single variance term on Ka labelled "Between-subject variability omega
  Ka" (102.5% CV) without a separate IOV row. The packaged model encodes
  this single 102.5% value as a single `etalka` variance, which captures
  the absorption variability the paper reports without distinguishing BSV
  from IOV; the published Table 3 does not provide enough information to
  separate the two terms, and no NONMEM control stream (`.mod` / `.lst`)
  is available with the paper.

- **IIV on ALAG not retained.** Jeon 2012 Methods describes BSV on ALAG
  among the screened random effects, but the final-model Table 3 omits an
  `omega_ALAG` row; the packaged model therefore treats ALAG as a typical-
  value-only parameter (no `etalalag`).

- **Sampling grid matches the paper.** Plasma sampling in the simulation
  uses the same 0.05 / 0.25 / 0.5 / 1 / 1.5 / 2 / 4 / 6 h post-dose grid
  the paper used (Methods "Blood sampling"), so the simulated NCA
  characteristics see the same Cmax / Tmax resolution as the observed
  data.

- **LLOQ.** The bioanalytical assay LLOQ for C3G was 0.2 ng/mL (Methods
  "Assay of the plasma concentration of C3G"). The packaged-model
  proportional residual error (21.3%) does not include an additive LLOQ
  floor; below-LLOQ behaviour is simulation-noise-only and should not be
  interpreted clinically.

- **No NONMEM control stream supplement.** The paper does not ship a
  `$THETA` / `$OMEGA` / `$SIGMA` block as a supplement. Final parameter
  values are taken directly from Table 3 of Jeon 2012.

- **Errata search.** A targeted PubMed search for `kjpp 2012 16 4 249
  erratum` and a Google Scholar search for `"Jeon 2012 cyanidin" erratum`
  returned no results as of the extraction date; no published correction
  is on file.
