---
title: "Ivermectin lung exposure minimal PBPK (Jermain 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ivermectin lung exposure minimal PBPK (Jermain 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4.5, dpi = 96
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Jermain et al. (2020) built a minimal physiologically-based pharmacokinetic
(mPBPK) model to ask whether oral ivermectin, repurposed against SARS-CoV-2,
reaches lung concentrations near its in vitro IC50 of 1750 ng/mL. The model has
a transit compartment and an absorption compartment feeding plasma, a lung
compartment that receives the whole cardiac output, and a lumped rest-of-body
("Other") compartment that receives a fraction of it. Both tissues are
perfusion-rate-limited.

```{r model}
mod <- readModelDb("Jermain_2020_ivermectin_pbpk")
ui <- rxode2::rxode(mod)
```

* Citation: `r ui$reference`
* Article: <https://doi.org/10.1016/j.xphs.2020.08.024>
* PubMed Central: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7473010/>

```{r description, echo = FALSE, results = "asis"}
cat(ui$description, "\n")
```

# Population

The human part of the model was not fitted to individual data. The authors took
the population PK model of a phase 1 study in 18 healthy volunteers given
0.2 mg/kg oral ivermectin (El-Tahtawy et al., reference 13 of the paper),
simulated its geometric-mean plasma profile for the study-average dose of 15 mg
without between-subject or residual variability, and fitted the five free mPBPK
parameters to that profile in Phoenix WinNonlin 8.2 (Methods, "Ivermectin
Pharmacokinetic Data" and "Parameters").

The lung partition coefficient came from a different species: it is the ratio
of lung to plasma AUC in Holstein calves given 0.2 mg/kg subcutaneous ivermectin
(Lifschitz et al., reference 12), assumed to be species-independent because
plasma protein binding is similarly high in cattle and humans. Physiological
volumes and cardiac output are population averages for adults (references 14
and 15). No covariates are carried.

# Source trace

| Quantity | Model name | Value | Source |
|---|---|---|---|
| Absorption rate constant | `lka` | 0.14 1/h | Table 1 (CV 2.52%) |
| Apparent plasma clearance CLp/F | `lcl` | 10.85 L/h | Table 1 (CV 3.78%) |
| Apparent Other partition coefficient Kp,other/F | `lkp_other` | 17.32 | Table 1 (CV 5.6%) |
| Fraction of cardiac output to Other | `fq_other` | 0.08 | Table 1 (CV 1.18%) |
| Transit rate constant | `lktr` | 0.36 1/h | Table 1 (CV 3.09%) |
| Apparent lung partition coefficient Kp,lung/F | `lkp_lung` | 2.68 (fixed) | Table 1, footnote a; calf AUC ratio (Methods) |
| Plasma volume | `lvc` | 3 L (fixed) | Table 1, footnote a (ref 15) |
| Lung volume | `lv_lung` | 1.3 L (fixed) | Table 1, footnote a (ref 14) |
| Remaining volume | `lv_other` | 65.7 L (fixed) | Table 1, footnote a |
| Cardiac output | `q_co` | 282 L/h (fixed) | Table 1, footnote a (ref 15) |
| Residual error, plasma and lung | `propSd`, `propSd_Clung` | 20% (fixed) | Methods, "Human Lung Exposure Simulation" |
| Transit and absorption ODEs | `depot`, `transit1` | -- | page 3576, equations for a0 and Absorption |
| Plasma ODE | `central` | -- | page 3576, plasma equation (full-width) |
| Lung and Other ODEs | `lung`, `other` | -- | page 3576, right column |
| Model schematic | -- | -- | Figure 1 |

# Typical-value simulation of the fitted 15 mg profile (Figure 2)

Figure 2 of the paper overlays the mPBPK fit on the simulated geometric-mean
plasma profile after 15 mg. The published points are not tabulated, so only
the model curve is reproduced here.

```{r fig2}
mod_typ <- rxode2::zeroRe(mod)

# The model has two error endpoints (plasma and lung), so each observation row
# names its endpoint with dvid; both Cc and Clung come back as columns.
make_events <- function(dose, times, n = 1) {
  dplyr::bind_rows(
    data.frame(id = seq_len(n), time = 0, amt = dose, evid = 1L,
               cmt = "depot", dvid = NA_integer_),
    expand.grid(id = seq_len(n), time = times) |>
      dplyr::mutate(amt = 0, evid = 0L, cmt = NA_character_, dvid = 1L)
  ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

sim15 <- as.data.frame(rxode2::rxSolve(mod_typ, make_events(15, seq(0, 72, by = 0.1))))

ggplot(sim15, aes(time, Cc)) +
  geom_line(colour = "steelblue") +
  labs(x = "Time after dose (h)", y = "Plasma ivermectin (ng/mL)",
       title = "Replicates Figure 2 of Jermain 2020 (model curve)",
       caption = "Single 15 mg oral dose, typical values.")
```

# Lung and plasma exposure after 12, 30 and 120 mg (Figure 3, Figure S1)

The authors simulated 1000 subjects per dose on a 0.1 h grid with an assumed
20% residual error and no between-subject variability, and plotted the median
and 95% interval. A cohort of 200 per dose is used here.

```{r fig3}
rxode2::rxSetSeed(20200904)
doses <- c(12, 30, 120)
times <- seq(0, 72, by = 0.1)

sim_sto <- dplyr::bind_rows(lapply(doses, function(d) {
  as.data.frame(rxode2::rxSolve(mod, make_events(d, times, n = 200))) |>
    dplyr::mutate(dose = d)
}))

band <- sim_sto |>
  tidyr::pivot_longer(c(Cc, Clung), names_to = "matrix", values_to = "conc") |>
  dplyr::mutate(matrix = ifelse(matrix == "Cc", "Plasma", "Lung")) |>
  dplyr::group_by(dose, matrix, time) |>
  dplyr::summarise(
    median = stats::median(conc),
    lo = stats::quantile(conc, 0.025),
    hi = stats::quantile(conc, 0.975),
    .groups = "drop"
  )

ggplot(band, aes(time, median, colour = matrix, fill = matrix)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, colour = NA) +
  geom_line() +
  geom_hline(yintercept = 1750, linetype = "dashed") +
  facet_wrap(~ paste(dose, "mg"), scales = "free_y") +
  labs(x = "Time after dose (h)", y = "Ivermectin (ng/mL)",
       colour = NULL, fill = NULL,
       title = "Replicates Figure 3 (120 mg) and Figure S1 (12, 30 mg)",
       caption = "Median and 95% interval of 200 simulated subjects; dashed line = in vitro IC50.")
```

The stochastic median tracks the typical-value profile, since the only
variability in the model is residual error:

```{r median-vs-typical}
sim120_typ <- as.data.frame(rxode2::rxSolve(mod_typ, make_events(120, times)))
med120 <- band |>
  dplyr::filter(dose == 120) |>
  dplyr::group_by(matrix) |>
  dplyr::summarise(cmax_median = max(median), .groups = "drop")
med120

stopifnot(
  abs(med120$cmax_median[med120$matrix == "Plasma"] / max(sim120_typ$Cc) - 1) < 0.1,
  abs(med120$cmax_median[med120$matrix == "Lung"] / max(sim120_typ$Clung) - 1) < 0.1
)
```

# PKNCA validation

NCA is run on the typical-value profiles for each dose, separately for plasma
and lung, over a long enough window (0-2000 h) that the extrapolated AUC is
negligible.

```{r pknca}
nca_times <- sort(unique(c(seq(0, 24, by = 0.1), seq(25, 2000, by = 5))))
sim_nca <- dplyr::bind_rows(lapply(doses, function(d) {
  as.data.frame(rxode2::rxSolve(mod_typ, make_events(d, nca_times))) |>
    dplyr::mutate(id = 1L, treatment = paste(d, "mg"), dose = d)
}))

run_nca <- function(conc_col) {
  conc_df <- sim_nca |>
    dplyr::transmute(id, treatment, time, conc = .data[[conc_col]]) |>
    dplyr::filter(!is.na(conc))
  dose_df <- sim_nca |>
    dplyr::distinct(id, treatment, dose) |>
    dplyr::mutate(time = 0)
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(conc_df, conc ~ time | treatment + id),
    PKNCA::PKNCAdose(dose_df, dose ~ time | treatment + id),
    intervals = data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                           aucinf.obs = TRUE, half.life = TRUE)
  ))
  as.data.frame(res)
}

nca_plasma <- run_nca("Cc") |> dplyr::mutate(matrix = "Plasma")
nca_lung <- run_nca("Clung") |> dplyr::mutate(matrix = "Lung")

nca_all <- dplyr::bind_rows(nca_plasma, nca_lung) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  dplyr::select(matrix, treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

nca_all |>
  dplyr::rename(
    "Matrix" = matrix, "Dose" = treatment, "Cmax (ng/mL)" = cmax,
    "Tmax (h)" = tmax, "AUC0-inf (h*ng/mL)" = aucinf.obs, "t1/2 (h)" = half.life
  ) |>
  knitr::kable(digits = 2, caption = "Typical-value NCA of plasma and lung.")
```

Two exact identities follow from the model structure and check that the ODEs
were transcribed with the right flows. First, dose over AUC in plasma recovers
CLp/F. Second, because the lung is perfused by the full cardiac output in and
out, its AUC is exactly Kp,lung/F times the plasma AUC; that ratio is how the
authors defined Kp,lung/F from the calf data.

```{r identities}
auc_p <- nca_plasma |> dplyr::filter(PPTESTCD == "aucinf.obs") |> dplyr::arrange(treatment)
auc_l <- nca_lung |> dplyr::filter(PPTESTCD == "aucinf.obs") |> dplyr::arrange(treatment)
dose_by_trt <- as.numeric(sub(" mg", "", auc_p$treatment))

cl_recovered <- 1000 * dose_by_trt / auc_p$PPORRES   # mg / (h*ng/mL) -> L/h
ratio_lung <- auc_l$PPORRES / auc_p$PPORRES
data.frame(treatment = auc_p$treatment, cl_recovered, ratio_lung)

stopifnot(
  all(abs(cl_recovered / 10.85 - 1) < 0.01),
  all(abs(ratio_lung / 2.68 - 1) < 0.01)
)
```

## Comparison against the published 120 mg peak

The paper reports the maximum simulated concentration after 120 mg as 288 ng/mL
in plasma and 772 ng/mL in lung, both at 5.1 h (Results, "Human Lung Exposure
Simulation"). Those values are read from the median of the authors' stochastic
simulation, which for a model whose only variability is residual error sits on
the typical-value profile.

```{r nca-compare}
simulated <- nca_all |>
  dplyr::filter(treatment == "120 mg") |>
  dplyr::transmute(group = matrix, cmax, tmax)

published <- data.frame(
  group = c("Plasma", "Lung"),
  cmax = c(288, 772),
  tmax = c(5.1, 5.1)
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = simulated,
  reference = published,
  by = "group",
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Simulated typical-value 120 mg peak versus Jermain 2020. * marks a difference greater than 20%.")

typ_cmax <- setNames(simulated$cmax, simulated$group)
stopifnot(
  abs(typ_cmax[["Plasma"]] / 288 - 1) < 0.05,
  abs(typ_cmax[["Lung"]] / 772 - 1) < 0.05
)
```

The simulated peaks agree with the paper to within about 1%, and the simulated
Tmax (about 4.8 h) is within 0.3 h of the published 5.1 h; the published value
is the peak of a median over noisy stochastic profiles, which shifts it
slightly. Neither dose reaches the 1750 ng/mL in vitro IC50 in lung, matching
the paper's conclusion.

# Assumptions and deviations

* **`Ka*Dose` read as `Ka*Absorption`.** The printed absorption and plasma
  equations (page 3576) write the absorption flux as `Ka*Dose`. Taken literally
  that is a constant input that never stops and drives the Absorption amount
  negative. Figure 1 draws the flux from the Absorption compartment to Plasma
  with rate Ka, and the text calls Ka a first-order absorption rate constant,
  so the flux is encoded as `ka * transit1` (the Absorption amount). With that
  reading the model reproduces the published 120 mg plasma and lung peaks to
  within about 1%.
* **Compartment naming.** The paper's transit compartment `a0` (which receives
  the dose) is `depot`, and its Absorption compartment is `transit1`, so the
  chain reads depot --ktr--> transit1 --ka--> central.
* **Residual error.** The paper used an "assumed 20% RUV" of unstated form for
  its lung-exposure simulations and does not say whether it applied the same
  error to lung. It is encoded as a fixed 20% proportional error on both the
  plasma and the lung output. It was not estimated; the fit itself was to a
  noise-free profile.
* **No between-subject variability.** None was estimated or used by the
  authors, so none is encoded; stochastic simulations vary only by residual
  error.
* **Apparent (F-scaled) parameters.** CLp/F, Kp,lung/F and Kp,other/F are used
  exactly as tabulated. Bioavailability is not separated out, so the dose is
  given into `depot` with F = 1.
* **Cardiac-output topology.** As printed, plasma loses flow to lung at the
  full cardiac output and to the Other compartment at `Fraction * Qco` in
  parallel, so the total flow out of plasma is `(1 + Fraction) * Qco`. That is
  the published equation set and it is reproduced as written.
* **Figure 2 data.** The simulated geometric-mean plasma concentrations the
  model was fitted to are shown only graphically; they were not digitised, and
  the Figure 2 replicate shows the model curve only.
* **Supplement.** The supplementary material (calf lung and plasma data,
  Table S1, and the 12 and 30 mg simulations, Figure S1) holds no model
  parameters; the 12 and 30 mg panels above are simulated from the Table 1
  parameters.
