---
title: "TDI01 (Ji 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TDI01 (Ji 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Ji_2025_TDI01"))$reference`
* Description: `r rxode2::rxode(readModelDb("Ji_2025_TDI01"))$description`
* Article: <https://doi.org/10.3389/fphar.2025.1477607>

TDI01 is a selective Rho-associated coiled-coil-containing protein kinase 2
(ROCK2) inhibitor developed for acute lung injury / acute respiratory distress
syndrome. Ji and Cui fitted a population PK model to a phase 1 single- and
multiple-ascending-dose study, then used it to simulate two candidate 7-day
regimens.

The defining feature of the data is a **double-peak** concentration-time
profile after a single oral dose, which the authors attributed to
hepato-enteral (enterohepatic) circulation. Their structural model therefore
carries a gallbladder compartment that fills continuously from the central
compartment and discharges back into the absorption compartment during a
2 h window, gated by a time switch the paper calls `FLAG`.

There is no supplementary material and no published erratum for this article
(checked 2026-08-14); the model specification is contained entirely in the
paper's Table 1 and Equations 1-5.

# Population

The model was fitted to 776 quantifiable plasma concentrations from
**39 healthy Chinese volunteers** in a single phase 1 study, sponsored by
Beijing Tide Pharmaceutical Co., Ltd. and conducted at the Institute of
Clinical Pharmacology, Peking University First Hospital. Single-dose arms
received 400, 800, or 1,200 mg orally; multiple-dose arms received 200 mg QD
or 400 mg QD for 7 days. Single-dose arms were sampled pre-dose and at 1, 2,
3, 3.5, 4, 5, 6, 8, 10, 12, 24, 36, 48, and 72 h; multiple-dose arms were
sampled out to 216 h with a dense schedule bracketing the day-7 dose.

Ten of 786 observations (1.3%) were below the limit of quantification; because
that is under the paper's 10% threshold, they were excluded rather than
handled by a likelihood-based method. Observations with `|CWRES| > 5` were
treated as outliers during model building and re-introduced for a sensitivity
analysis.

**The paper publishes no baseline demographic table.** Age, body weight, sex,
and BMI distributions are not reported; the only characterisation given is
that the cohort had a "homogeneous demographic profile" (Discussion,
Limitations). Consequently no covariate model was developed, and the
`population` metadata for this model deliberately omits those fields rather
than guessing them. The virtual cohort below therefore varies only the
model's own random effects.

```{r population}
pop <- rxode2::rxode(readModelDb("Ji_2025_TDI01"))$population
str(pop, max.level = 1)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Ji_2025_TDI01.R`. The table below collects
them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL/F) | 95 L/h | Table 1, "CL/F (L/h)", Final model (RSE 10%) |
| `lvc` (Vc/F) | 1,400 L | Table 1, "Vc/F (L)", Final model (RSE 8%) |
| `lka` (ka/F) | 0.345 1/h | Table 1, "ka/F (1/h)", Final model (RSE 10%) |
| `lkbm` (k2G/F) | 0.023 1/h, FIXED | Table 1, "k2G/F (1/h)" = 0.023 (FIX) |
| `lkehc` (kG1/F) | 2 1/h, FIXED | Table 1, "kG1/F (1/h)" = 2 (FIX) |
| `ltgb` (FLAG onset) | 8 h post-dose, FIXED | Results: "the starting and ending time of hepato-enteral circulation was set as 8-10 h" |
| `ldgb` (FLAG duration) | 2 h, FIXED | Results, same sentence (10 h - 8 h) |
| `etalcl` | 0.198051 = log(1 + 0.468^2) | Table 1, "IIV_CL/F" = 46.8% CV; Methods Equation 1 (exponential IIV) |
| `etalvc` | 0.098654 = log(1 + 0.322^2) | Table 1, "IIV_V/F" = 32.2% CV; Methods Equation 1 |
| `etalka` | 0.201903 = log(1 + 0.473^2) | Table 1, "IIV_k12" = 47.3% CV; Methods Equation 1 (see note below) |
| `propSd` | 0.342 | Table 1, "Prop.error (%)" = 34.2% (RSE 3%); Methods Equation 2 |
| `addSd` | 1.77 ng/mL | Table 1, "Add.error" = 1.77 (RSE 47%); Methods Equation 2 |
| `kel <- cl / vc` | n/a | Line preceding Equation 3: "K20 = CL/Vc" |
| `d/dt(depot)` | n/a | Equation 3: `DADT(1) = FLAG*KG1*A(2) - Ka*A` |
| `d/dt(central)` | n/a | Equation 4: `DADT(2) = Ka*A - K2G*A(1) - K20*A(1)` |
| `d/dt(gallbladder)` | n/a | Equation 5: `DADT(3) = K2G*A(1) - FLAG*KG1*A(2)` |
| `Cc <- central / vc * 1000` | n/a | Unit conversion mg/L -> ng/mL; concentration scale set by Table 3 (ng/mL) and Add.error = 1.77 |

Two transcription decisions in that table deserve their reasoning spelled out.

**Compartment indexing.** Equations 3-5 use `DADT(1)` / `DADT(2)` / `DADT(3)`
on the left-hand side but `A` / `A(1)` / `A(2)` on the right, and the legend
states that "A, A (1) and A (2) represent absorption, central and gallbladder
compartments, respectively". Reading the right-hand sides under that legend,
the three equations are internally consistent and map unambiguously onto
`depot` / `central` / `gallbladder`: every gallbladder loss term is exactly
the matching depot gain term, and the central loss to bile is exactly the
gallbladder gain. Mass balance therefore closes, so the index offset between
the `DADT(n)` and `A(n)` notations is a typographic artifact rather than a
genuine ambiguity.

**`IIV_k12`.** Table 1's third variance component is labelled `IIV_k12`, a
name that appears nowhere in Equations 3-5. It is the IIV on `ka`: `k12` is
the NONMEM micro-constant for transfer from compartment 1 (depot) to
compartment 2 (central), which is `ka`; and `ka` is the only estimated
structural parameter left once `IIV_CL/F` and `IIV_V/F` are assigned, because
`k2G` and `kG1` are both FIX and so cannot carry a variance.

# Virtual cohort

Original observed data are not publicly available. The figures below use
virtual cohorts of 200 subjects per arm (the per-arm cap for this repository),
varying only the model's own between-subject random effects on CL/F, Vc/F, and
`ka` -- the paper reports no covariates and no demographic distributions to
sample from.

```{r cohort}
set.seed(20250304)

n_arm <- 200L

# One arm as a self-contained event table. `id_offset` keeps IDs disjoint
# across arms; duplicate IDs across arms would silently collapse into single
# subjects that receive the summed dose.
make_arm <- function(amt, ii, addl, label, obs_times, id_offset = 0L) {
  ev <- rxode2::et(amt = amt, ii = ii, addl = addl, cmt = "depot") |>
    rxode2::et(obs_times, cmt = "central") |>
    rxode2::et(id = id_offset + seq_len(n_arm))
  as.data.frame(ev) |>
    dplyr::mutate(regimen = label)
}

# --- single ascending doses (paper Figure 4) -------------------------------
sd_times <- sort(unique(c(seq(0, 14, by = 0.25), seq(14, 72, by = 0.5))))

events_sd <- dplyr::bind_rows(
  make_arm(400,  ii = 0, addl = 0, "400 mg single",   sd_times, id_offset = 0L),
  make_arm(800,  ii = 0, addl = 0, "800 mg single",   sd_times, id_offset = 1000L),
  make_arm(1200, ii = 0, addl = 0, "1,200 mg single", sd_times, id_offset = 2000L)
)
stopifnot(!anyDuplicated(events_sd[, c("id", "time", "evid")]))

# --- multiple ascending doses + the two simulated regimens (Figures 5, 6) ---
# Dense sampling brackets the day-7 dose (144 h) so the steady-state interval
# is well resolved for NCA.
md_times <- sort(unique(c(seq(0, 24, by = 0.5), seq(24, 144, by = 2),
                          seq(144, 168, by = 0.25), seq(168, 216, by = 1))))

events_md <- dplyr::bind_rows(
  make_arm(200, ii = 24, addl =  6, "200 mg QD, 7 d",  md_times, id_offset = 0L),
  make_arm(400, ii = 24, addl =  6, "400 mg QD, 7 d",  md_times, id_offset = 1000L),
  make_arm(200, ii = 12, addl = 13, "200 mg BID, 7 d", md_times, id_offset = 2000L)
)
stopifnot(!anyDuplicated(events_md[, c("id", "time", "evid")]))

c(single_dose_rows = nrow(events_sd), multiple_dose_rows = nrow(events_md))
```

# Simulation

```{r simulate}
mod <- readModelDb("Ji_2025_TDI01")

sim_sd <- rxode2::rxSolve(mod, events = events_sd, keep = "regimen") |>
  as.data.frame()
sim_md <- rxode2::rxSolve(mod, events = events_md, keep = "regimen") |>
  as.data.frame()

# Typical-value (deterministic) profiles: supply zero eta columns and switch
# off the sampler with omega = NA. This is preferred over zeroRe(), which
# mutates shared model state.
etas0 <- data.frame(id = 1L, etalcl = 0, etalvc = 0, etalka = 0)

solve_typical <- function(amt, ii, addl, obs_times, pars = NULL) {
  ev <- rxode2::et(amt = amt, ii = ii, addl = addl, cmt = "depot") |>
    rxode2::et(obs_times, cmt = "central")
  rxode2::rxSolve(mod, ev, etas0, omega = NA, params = pars,
                  atol = 1e-10, rtol = 1e-10) |>
    as.data.frame()
}
```

# Replicate published figures

## Figure 1 -- dose-normalised profiles

Figure 1 of Ji 2025 overlays dose-normalised concentration-time profiles from
every arm. In the observed data the low-dose arms sit **above** the high-dose
arms, which is the observation the authors cite as evidence of a "dosage
effect" on bioavailability. The packaged model is strictly linear in dose
(`f(depot)` is unity at every dose and every term is first-order), so its
dose-normalised profiles superimpose exactly. That superposition is the
cleanest available demonstration that the dose effect described in the text
was never quantified into a parameter -- see *Assumptions and deviations*.

```{r figure-1}
# Replicates Figure 1 of Ji 2025: dose-normalised concentration-time profiles.
dose_of <- c("400 mg single" = 400, "800 mg single" = 800,
             "1,200 mg single" = 1200)

sim_sd |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(Cc_norm = Cc / dose_of[as.character(regimen)]) |>
  dplyr::group_by(time, regimen) |>
  dplyr::summarise(Q50 = median(Cc_norm), .groups = "drop") |>
  ggplot(aes(time, Q50, colour = regimen)) +
  geom_line(linewidth = 0.9) +
  labs(x = "Time (h)", y = "Dose-normalised Cc (ng/mL per mg)",
       colour = "Regimen",
       title = "Figure 1 -- dose-normalised single-dose profiles",
       caption = paste("Replicates Figure 1 of Ji 2025. The three curves",
                       "superimpose because the packaged model is linear in",
                       "dose; the paper's observed curves do not.")) +
  theme(legend.position = "bottom")
```

## Figures 4 and 5 -- visual predictive checks

```{r figure-4}
# Replicates Figure 4 of Ji 2025: VPC of single ascending doses.
vpc <- function(d) {
  d |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::group_by(time, regimen) |>
    dplyr::summarise(Q025 = quantile(Cc, 0.025), Q50 = median(Cc),
                     Q975 = quantile(Cc, 0.975), .groups = "drop")
}

vpc(sim_sd) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q025, ymax = Q975), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  facet_wrap(~regimen) +
  labs(x = "Time (h)", y = "Cc (ng/mL)",
       title = "Figure 4 -- VPC, single ascending doses",
       caption = paste("Replicates Figure 4 of Ji 2025 (arithmetic scale).",
                       "Median and 95% prediction interval, 200 subjects/arm."))
```

```{r figure-4-log}
# Figure 4B of Ji 2025: the same VPC on a logarithmic concentration scale.
vpc(sim_sd) |>
  dplyr::filter(time > 0) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q025, ymax = Q975), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  facet_wrap(~regimen) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Cc (ng/mL, log scale)",
       title = "Figure 4B -- VPC, single ascending doses (log scale)",
       caption = "Replicates Figure 4B of Ji 2025.")
```

```{r figure-5}
# Replicates Figure 5 of Ji 2025: VPC of multiple ascending doses (the two
# clinical multiple-dose arms, 200 mg QD and 400 mg QD for 7 days).
vpc(sim_md) |>
  dplyr::filter(regimen %in% c("200 mg QD, 7 d", "400 mg QD, 7 d")) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q025, ymax = Q975), alpha = 0.25, fill = "darkorange") +
  geom_line(colour = "darkorange3", linewidth = 0.8) +
  facet_wrap(~regimen, ncol = 1) +
  labs(x = "Time (h)", y = "Cc (ng/mL)",
       title = "Figure 5 -- VPC, multiple ascending doses",
       caption = paste("Replicates Figure 5 of Ji 2025. Doses at 0-144 h;",
                       "sampling continues to 216 h."))
```

## Figure 6 -- the two simulated candidate regimens

```{r figure-6}
# Replicates Figure 6 of Ji 2025: simulated profiles for the two designed
# regimens over the 7-day treatment course.
vpc(sim_md) |>
  dplyr::filter(regimen %in% c("400 mg QD, 7 d", "200 mg BID, 7 d")) |>
  ggplot(aes(time, Q50, colour = regimen, fill = regimen)) +
  geom_ribbon(aes(ymin = Q025, ymax = Q975), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 1391, linetype = "dashed") +
  annotate("text", x = 20, y = 1391, vjust = -0.5, size = 3,
           label = "observed Cmax at 1,200 mg single dose = 1,391 ng/mL") +
  labs(x = "Time (h)", y = "Cc (ng/mL)", colour = "Regimen", fill = "Regimen",
       title = "Figure 6 -- simulated candidate regimens",
       caption = paste("Replicates Figure 6 of Ji 2025. The dashed line is the",
                       "safety benchmark the paper compares against.")) +
  theme(legend.position = "bottom")
```

The paper's safety argument is that concentrations under both 7-day regimens
stay below the 1,391 ng/mL Cmax observed at the highest single dose. The
packaged model reproduces that conclusion, and with a wide margin.

# Structural checks

Three quantitative gates exercise parts of the model that a visual profile
check cannot. The first two are exact identities; the third confirms the
enterohepatic loop is actually live rather than silently inert.

## Closed-form steady-state mass balance

Because every molecule that enters the gallbladder is eventually released back
into the depot and cleared only through CL/F, the steady-state exposure over a
24 h period must equal the daily dose divided by apparent clearance,
*independently of the recirculation parameters*:

$$\mathrm{AUC}_{0-24,ss} = \frac{D_{\text{daily}}}{\mathrm{CL}/F}$$

This is a strong gate: it fails if the recirculation is leaking mass, if the
`FLAG` window never reopens, or if the mg/L-to-ng/mL scaling is wrong.

```{r gate-mass-balance}
auc_trap <- function(d) sum(diff(d$time) * (head(d$Cc, -1) + tail(d$Cc, -1)) / 2)

ss_window <- seq(144, 168, by = 0.05)

mb <- dplyr::bind_rows(
  lapply(
    list(list("400 mg QD, 7 d", 400, 24, 6), list("200 mg BID, 7 d", 200, 12, 13)),
    function(r) {
      d <- solve_typical(r[[2]], r[[3]], r[[4]], ss_window) |>
        dplyr::filter(time >= 144, time <= 168)
      data.frame(regimen = r[[1]],
                 simulated = auc_trap(d),
                 closed_form = (24 / r[[3]]) * r[[2]] / 95 * 1000)
    }
  )
) |>
  dplyr::mutate(pct_diff = 100 * (simulated - closed_form) / closed_form)

mb |>
  dplyr::rename("Regimen" = regimen,
                "Simulated AUC0-24,ss (ng*h/mL)" = simulated,
                "Dose_daily / (CL/F) (ng*h/mL)"  = closed_form,
                "% diff" = pct_diff) |>
  knitr::kable(digits = c(0, 1, 1, 3),
               caption = "Steady-state AUC against the closed-form identity.")

stopifnot(all(abs(mb$pct_diff) < 0.5))
```

Both regimens agree with the closed form to better than 0.5%, and they agree
with **each other** -- which is expected, because 400 mg QD and 200 mg BID
deliver the same 400 mg daily dose. Hold on to that fact; it matters when the
paper's own Table 3 is compared below.

## Analytic Tmax and terminal half-life

Before the `FLAG` window opens at 8 h the gallbladder is a pure sink, so the
single-dose system is exactly a one-compartment oral model with total central
exit rate $k = k_{el} + k_{2G}$. Tmax and the terminal slope therefore have
closed forms, and Tmax (5.25 h) falls before the window opens, so the
recirculation cannot perturb it.

```{r gate-tmax}
cl <- 95; vc <- 1400; ka <- 0.345; k2g <- 0.023
k  <- cl / vc + k2g

tmax_analytic <- log(ka / k) / (ka - k)
thalf_analytic <- log(2) / k

d1200 <- solve_typical(1200, 0, 0, seq(0, 72, by = 0.01))
tmax_sim <- d1200$time[which.max(d1200$Cc)]

tail_fit <- d1200 |> dplyr::filter(time >= 40, time <= 70)
thalf_sim <- log(2) / -coef(lm(log(Cc) ~ time, data = tail_fit))[["time"]]

data.frame(
  quantity  = c("Tmax (h)", "Terminal half-life (h)"),
  analytic  = c(tmax_analytic, thalf_analytic),
  simulated = c(tmax_sim, thalf_sim)
) |>
  dplyr::mutate("% diff" = 100 * (simulated - analytic) / analytic) |>
  dplyr::rename("Quantity" = quantity, "Analytic" = analytic,
                "Simulated" = simulated) |>
  knitr::kable(digits = 3,
               caption = "Single-dose Tmax and terminal half-life vs. closed form.")

stopifnot(abs(tmax_sim - tmax_analytic) < 0.05,
          abs(thalf_sim - thalf_analytic) / thalf_analytic < 0.01)
```

## The enterohepatic loop is live

A named intermediate that gates an ODE term can silently evaluate to zero, in
which case the whole recirculation would vanish without any error. This gate
ablates biliary uptake (`k2G` set near zero) and confirms that the gallbladder
state, and the concentration trajectory, both respond.

```{r gate-ehc-live}
d_ehc  <- solve_typical(1200, 0, 0, seq(0, 48, by = 0.05))
d_none <- solve_typical(1200, 0, 0, seq(0, 48, by = 0.05),
                        pars = c(lkbm = log(1e-12)))

gb_peak  <- max(d_ehc$gallbladder[d_ehc$time <= 8])
gb_after <- d_ehc$gallbladder[which.min(abs(d_ehc$time - 10))]

data.frame(
  quantity = c("Gallbladder amount at 8 h, FLAG onset (mg)",
               "Gallbladder amount at 10 h, FLAG close (mg)",
               "Fraction of the gallbladder emptied in the window",
               "Gallbladder peak with k2G ablated (mg)"),
  value = c(gb_peak, gb_after, 1 - gb_after / gb_peak,
            max(d_none$gallbladder))
) |>
  dplyr::rename("Quantity" = quantity, "Value" = value) |>
  knitr::kable(digits = 4, caption = "Ablation check on the enterohepatic loop.")

# The loop must move a material amount of drug, and must vanish when ablated.
stopifnot(gb_peak > 100,                       # >100 mg of the 1,200 mg dose
          1 - gb_after / gb_peak > 0.9,        # window empties >90% of it
          max(d_none$gallbladder) < 1e-3)      # ablation kills the loop
```

The recirculation moves `r round(gb_peak)` mg of the 1,200 mg dose and
discharges over 90% of it during the 2 h window, and ablating `k2G` removes
the loop entirely. The mechanism is therefore live.

## What the recirculation does *not* do

With the published parameter values the recirculation does **not** produce a
visible second peak. The recycled bolus re-enters the *depot* and must then be
absorbed at `ka` = 0.345 1/h (2.0 h half-life) into a central compartment that
is still holding roughly 700 mg and declining with a 7.6 h half-life, so the
returning mass changes the curvature without ever reversing the slope.

```{r ehc-signature}
sig <- d_ehc |>
  dplyr::select(time, Cc_ehc = Cc) |>
  dplyr::inner_join(d_none |> dplyr::select(time, Cc_none = Cc), by = "time") |>
  dplyr::mutate(ratio = Cc_ehc / Cc_none)

n_peaks <- function(x) sum(diff(sign(diff(x))) == -2)
at <- function(tt) sig$ratio[which.min(abs(sig$time - tt))]

# The recirculation's signature is a LOCAL minimum in the ratio at the FLAG
# onset followed by a rise as recycled drug is reabsorbed. It is not the
# global minimum over the whole profile: after a single dose the window never
# reopens, so biliary uptake continuing past 10 h is permanently stranded and
# the ratio resumes falling (quantified in the next chunk).
local_win <- sig |> dplyr::filter(time >= 4, time <= 14)
t_local_min <- local_win$time[which.min(local_win$ratio)]

cat(sprintf("local maxima in Cc: with EHC = %d, with k2G ablated = %d\n",
            n_peaks(d_ehc$Cc[d_ehc$time > 0]),
            n_peaks(d_none$Cc[d_none$time > 0])))
cat(sprintf("ratio on [4, 14] h is minimised at t = %.2f h (FLAG onset is 8 h)\n",
            t_local_min))
cat(sprintf("ratio: %.4f at 8.5 h -> %.4f at 12 h (rise = recycled drug reabsorbed)\n",
            at(8.5), at(12)))

stopifnot(t_local_min >= 8, t_local_min <= 9,   # local min sits at the onset
          at(12) > at(8.5),                     # then rises as drug reabsorbs
          n_peaks(d_ehc$Cc[d_ehc$time > 0]) == 1L)  # but never a second peak

ggplot(sig |> dplyr::filter(time > 0), aes(time, ratio)) +
  geom_vline(xintercept = c(8, 10), linetype = "dotted") +
  geom_line(linewidth = 0.8, colour = "purple4") +
  labs(x = "Time (h)", y = "Cc with EHC / Cc with k2G ablated",
       title = "Signature of the recirculation on a single 1,200 mg dose",
       caption = paste("Dotted lines bracket the 8-10 h FLAG window. The ratio",
                       "dips to a local minimum at the window onset and rises",
                       "as recycled drug is reabsorbed, then falls again as",
                       "post-window biliary uptake is stranded."))
```

After a single dose the `FLAG` window opens exactly once, so any drug taken up
into bile after 10 h stays there permanently. That is a real property of the
published model rather than an artifact of this implementation, and it means
single-dose AUC falls short of `D/(CL/F)` even though steady-state AUC matches
it exactly (as the mass-balance gate above showed).

```{r gate-single-dose-stranding}
strand <- data.frame(
  quantity = c("Gallbladder amount at 48 h (mg, of a 1,200 mg dose)",
               "Fraction of the dose permanently stranded",
               "AUC0-48 with EHC / AUC0-48 with k2G ablated"),
  value = c(
    tail(d_ehc$gallbladder, 1),
    tail(d_ehc$gallbladder, 1) / 1200,
    auc_trap(d_ehc) / auc_trap(d_none)
  )
)

strand |>
  dplyr::rename("Quantity" = quantity, "Value" = value) |>
  knitr::kable(digits = 4,
               caption = "Single-dose mass stranded by the one-shot FLAG window.")

# Steady state restores mass balance; a single dose does not.
stopifnot(tail(d_ehc$gallbladder, 1) / 1200 > 0.1,
          auc_trap(d_ehc) / auc_trap(d_none) < 1)
```

The ratio dips to a local minimum at the window onset and rises afterwards as
recycled drug is reabsorbed -- a clean quantitative signature of the
recirculation -- but the concentration trace itself has a single maximum. The
published model, as parameterised in Table 1, reproduces the *mechanism* the
authors describe without reproducing the *double peak* that motivated it.

This is worth stating plainly because it is a limitation of the source rather
than of the packaging: `k2G` = 0.023 1/h diverts only about 17% of the dose
into bile over the first 8 h, and the returning mass is delivered to the depot
rather than directly to plasma, so it must clear a 2.0 h absorption
half-life before it can influence the concentration. A crisp second peak would
require either a larger `k2G` or a return route into the central compartment.
Ji 2025 itself notes "some bias in predicting the high concentrations"
(Discussion).

# PKNCA validation

```{r pknca-sd}
sim_nca_sd <- sim_sd |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, regimen)

# Guarantee a time = 0 row per (id, regimen); pre-dose Cc = 0 is correct for
# an extravascular model.
sim_nca_sd <- dplyr::bind_rows(
  sim_nca_sd,
  sim_nca_sd |> dplyr::distinct(id, regimen) |>
    dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, regimen, time, .keep_all = TRUE) |>
  dplyr::arrange(id, regimen, time)

conc_sd <- PKNCA::PKNCAconc(sim_nca_sd, Cc ~ time | regimen + id,
                            concu = "ng/mL", timeu = "h")

dose_sd <- events_sd |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, regimen)

dose_obj_sd <- PKNCA::PKNCAdose(dose_sd, amt ~ time | regimen + id,
                                doseu = "mg")

intervals_sd <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

nca_sd <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_sd, dose_obj_sd, intervals = intervals_sd)
)

summary(nca_sd)
```

The single-dose NCA confirms dose proportionality (the model is linear in
dose) and a terminal half-life matching the 7.6 h closed form. Ji 2025
publishes no single-dose NCA table, so there is nothing to compare these
against numerically; the paper's only quantitative NCA output is the
steady-state Table 3, handled next.

```{r pknca-ss}
tau_of <- c("200 mg QD, 7 d" = 24, "400 mg QD, 7 d" = 24, "200 mg BID, 7 d" = 12)

sim_nca_md <- sim_md |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, regimen)

conc_md <- PKNCA::PKNCAconc(sim_nca_md, Cc ~ time | regimen + id,
                            concu = "ng/mL", timeu = "h")

dose_md <- events_md |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, regimen)

dose_obj_md <- PKNCA::PKNCAdose(dose_md, amt ~ time | regimen + id,
                                doseu = "mg")

# Table 3 reports Cmax,7d and AUC0-24h,7d: the day-7 (144-168 h) window,
# which spans one dosing interval for QD and two for BID.
intervals_md <- data.frame(
  start = 144, end = 168,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, cav = TRUE, cmin = TRUE
)

nca_md <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_md, dose_obj_md, intervals = intervals_md)
)

summary(nca_md)
```

## Comparison against published NCA

Table 3 of Ji 2025 reports the median and 95% interval of steady-state Cmax
and AUC0-24h on day 7 for the two candidate regimens.

```{r nca-compare}
published <- tibble::tribble(
  ~regimen,           ~cmax,  ~auclast,
  "400 mg QD, 7 d",   409.11, 4939.64,
  "200 mg BID, 7 d",  466.18, 6078.02
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_md,
  reference     = published,
  by            = "regimen",
  units         = c(cmax = "ng/mL", auclast = "ng*h/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated vs. Ji 2025 Table 3. * differs from reference by >20%.",
  align   = c("l", "l", "r", "r", "r")
)

attr(cmp, "footnote")
```

**Every Cmax row is starred, and the 200 mg BID AUC row is starred.** This is
not a transcription error and it is not tuneable away; it is a genuine
internal inconsistency in the source, and it has a single root cause. Three
independent lines of evidence identify it.

```{r table3-diagnosis}
tab3 <- tibble::tribble(
  ~regimen,          ~daily_mg, ~auc_paper, ~cmax_paper,
  "400 mg QD, 7 d",        400,    4939.64,      409.11,
  "200 mg BID, 7 d",       400,    6078.02,      466.18
) |>
  dplyr::mutate(
    auc_closed_form = daily_mg / 95 * 1000,
    auc_ratio       = auc_paper / auc_closed_form
  )

tab3 |>
  dplyr::rename("Regimen" = regimen, "Daily dose (mg)" = daily_mg,
                "Table 3 AUC0-24,7d" = auc_paper, "Table 3 Cmax,7d" = cmax_paper,
                "Dose_daily/(CL/F)" = auc_closed_form,
                "Table 3 / closed form" = auc_ratio) |>
  knitr::kable(digits = c(0, 0, 2, 2, 1, 3),
               caption = "Ji 2025 Table 3 against the model's own closed form.")
```

1. **The two regimens deliver identical daily doses** (400 mg QD and
   200 mg BID are both 400 mg/day), so any model in which clearance is
   dose-independent must give them the *same* steady-state AUC0-24. Table 3
   reports them 23% apart.
2. **Both Table 3 AUC values exceed** `D_daily / (CL/F)` = 4,211 ng*h/mL, by
   17% and 44% respectively. No parameterisation of Equations 3-5 can produce
   a steady-state AUC above that ceiling, because CL/F is the only route out
   of the system.
3. **Table 3's BID Cmax exceeds its QD Cmax** (466 vs 409 ng/mL) even though
   the BID arm administers half as much drug per dose. With a 10.2 h central
   half-life the q12h accumulation ratio (1.79) does not come close to
   offsetting the halved dose (a linear model predicts a BID/QD Cmax ratio
   near 0.7-0.9; the packaged model gives 0.88).

All three are explained by a dose-dependent bioavailability that the paper
describes in prose but never quantifies: the Results call the final model a
"one-compartment model with dosage effect", and the Discussion reports that
dose-normalised exposure implied higher bioavailability at 200-400 mg than at
800-1,200 mg. No such parameter appears in Table 1 and no dose term appears in
Equations 3-5. The Monte Carlo simulations behind Table 3 evidently applied
one; the published model specification does not contain it, so the packaged
model cannot and does not reproduce Table 3.

The AUC0-24 rows are the more diagnostic comparison, and the 400 mg QD AUC
lands within 17% of the published median despite this. Cmax is the more
sensitive of the two to the missing dose effect.

# Assumptions and deviations

* **Unquantified "dosage effect" (the principal deviation).** The source
  describes its final model as a one-compartment model "with dosage effect"
  and reports that relative bioavailability was higher at 200-400 mg than at
  800-1,200 mg, adding that this "need[s] more clinical evidence to support
  this conclusion". Table 1 contains no dose-effect parameter and
  Equations 3-5 contain no dose term, so **none was invented**: `f(depot)` is
  unity at every dose and every disposition parameter remains apparent
  (CL/F, Vc/F, ka/F). The consequences are quantified above: dose-normalised
  profiles superimpose exactly (Figure 1), and Table 3 is not reproducible
  (see the diagnosis table). A user who wishes to explore a dose effect can
  set `f(depot)` externally per dose group.
* **`FLAG` gated on time after dose.** The paper states only that "the
  starting and ending time of hepato-enteral circulation was set as 8-10 h"
  and does not say whether `FLAG` keys on absolute time or time after dose.
  Time after dose (`tad(depot)`) was chosen because the authors simulate 7-day
  QD and BID regimens and attribute accumulation to the recirculation, which
  requires the window to reopen after each dose. The reading is confirmed by
  the mass-balance gate above: with time-after-dose gating, steady-state AUC
  matches `D_daily/(CL/F)` to within 0.5%, whereas absolute-time gating would
  fire the window once in seven days and permanently strand a large fraction
  of each dose in the gallbladder. For a single dose the two readings are
  identical, so no published figure discriminates them.
* **The one-shot window strands mass after a single dose.** Because `FLAG`
  opens only on the 8-10 h post-dose window, drug taken up into bile after
  10 h has no release route until the next dose. After a *single* dose there
  is no next dose, so roughly 17% of the administered amount remains in the
  gallbladder indefinitely and single-dose AUC falls about 15% short of
  `D/(CL/F)`. This is a property of the published `FLAG` design, not of this
  implementation: any faithful encoding of Equations 3-5 with a bounded window
  behaves the same way. Under repeated dosing the window reopens each day and
  mass balance is restored exactly, which the steady-state gate above confirms
  to within 0.07%. Users simulating single doses should be aware that the
  terminal tail is slightly under-predicted for this reason.
* **`FLAG` window bounds encoded as onset plus duration.** `ltgb` = 8 h and
  `ldgb` = 2 h reproduce the stated 8-10 h window; the 2 h duration is the
  arithmetic difference, not a separately published value. Both are wrapped
  in `fixed()` because the authors set rather than estimated them (Table 1
  gives them no estimate, RSE, or bootstrap interval).
* **IIV percentages read as CVs of a log-normal.** Methods Equation 1 is an
  exponential IIV model, so the Table 1 percentages convert to log-scale
  variances via `omega^2 = log(1 + CV^2)`. Some NONMEM workflows instead
  print `sqrt(omega^2) * 100` under a "CV%" heading; under that alternative
  reading the omegas would be 0.468 / 0.322 / 0.473 rather than 0.445 /
  0.314 / 0.449, a difference of about 5% in relative terms. The paper does
  not state which convention its percentages follow.
* **`IIV_k12` assigned to `ka`.** Table 1's third variance component is
  labelled with a micro-constant that does not appear in the published
  equations. It is assigned to `ka` for the reasons given in the *Source
  trace* section (NONMEM compartment-1-to-2 transfer is `ka`, and `ka` is the
  only estimated structural parameter left once CL/F and Vc/F are assigned,
  since `k2G` and `kG1` are FIX).
* **Residual-error roles taken from the equation, not the prose.** Methods
  Equation 2 is `Yij = Cij * (1 + eps1) + eps2`, making `eps1` proportional
  and `eps2` additive. The surrounding sentence labels them the other way
  round, contradicting its own equation. The equation governs, and the
  Table 1 magnitudes corroborate it: 34.2% is dimensionless while 1.77 is on
  the ng/mL scale.
* **Table 1's k2G / kG1 definitions are garbled; Figure 2 and the equations
  govern.** Table 1 describes `k2G` as the rate "between central compartment
  and drug disposal compartment" and `kG1` as the rate "between central
  compartment and gallbladder compartment", which are neither distinct nor
  consistent with each other. Figure 2's legend is unambiguous (`K2G`:
  central to gallbladder; `KG1`: gallbladder to absorption compartment) and
  matches Equations 3-5. The magnitudes corroborate this direction: slow
  biliary uptake (0.023 1/h) with near-instant release (2 1/h) is the same
  pattern as `Courlet_2023_cabamiquine`, whose source supplement uses the
  identical `k2g` / `kg1` notation and fixes them at 0.0039 and 15 1/h.
* **No demographic distributions are sampled.** The paper publishes no
  baseline demographic table and develops no covariate model, so the virtual
  cohort varies only the model's own random effects. `population` omits
  `age_range`, `weight_range`, and `sex_female_pct` rather than guessing them.
* **The observed 1,391 ng/mL benchmark is an individual observation.** The
  paper quotes it as "the Cmax of 1,391 ng/mL under the highest single dose
  (1,200 mg)". The packaged model's typical-value Cmax at 1,200 mg is about
  530 ng/mL, and 1,391 ng/mL sits above the 97.5th percentile of a
  200-subject virtual cohort. It is used above only as the safety threshold
  the paper draws in its Figure 6 comparison, not as an NCA reference value.
* **Parameters are final estimates.** Table 1's "Final model" column is
  accompanied by a 1,000-replicate non-parametric bootstrap whose medians and
  95% intervals bracket every estimate, confirming these are converged final
  values rather than initial estimates.
