---
title: "Peginterferon alfa-2b / Pegbing (Jian 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Peginterferon alfa-2b / Pegbing (Jian 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Jian_2025_peginterferon_alfa_2b"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1002/psp4.70104> (CPT Pharmacometrics Syst Pharmacol. 2025;14:2014-2025; PMC12706394, open access)

Pegbing is a 40 kDa Y-shaped branched-PEG peginterferon alfa-2b (Xiamen Amoytop
Biotech). Jian 2025 is the first population PK analysis of the molecule and the
first PegIFN-alpha model to combine target-mediated drug disposition (TMDD) with
a time-dependent downregulation of interferon-receptor synthesis, fitted jointly
to healthy volunteers and chronic hepatitis B (CHB) patients.

Note that this molecule shares its INN with the 12 kDa linear-PEG product
(PegIntron) modelled in `Gupta_2006_peginterferon_alfa_2b`; the two are
structurally distinct and are **not** interchangeable.

# Population

```{r population, echo = FALSE}
pop <- ui$population
tibble::tibble(Field = names(pop), Value = vapply(pop, paste, character(1), collapse = "; ")) |>
  knitr::kable(caption = "Model `population` metadata (Jian 2025 Table 1).")
```

The modelling dataset pooled two trials (Jian 2025 Table 1): a Phase I
single-ascending-dose study in 28 healthy volunteers (single subcutaneous doses
of 45, 90, 180 or 270 ug; 428 observations) and the dense-PK subset of the
Phase II multiple-ascending-dose trial NCT01143662 in 39 CHB patients (90, 135
or 180 ug subcutaneously once weekly for 48 weeks). Together these contributed
67 subjects and 1013 observations. A separate sparse-PK subset of 115 CHB
patients (464 observations) was held out for external pcVPC validation and did
not inform the parameter estimates.

Both cohorts were young (median age 30 and 25 years), of similar body weight
(median 59.0 and 60.0 kg) and had normal renal function (Cockcroft-Gault CrCL
medians 108.95 and 120.49 mL/min). They differed sharply in liver-function
markers -- median ALT 20 U/L in healthy subjects versus 130 U/L in CHB patients
-- so any liver marker screened as a covariate is largely collinear with the
health-status indicator that the final model retained.

The same information is available programmatically via
`readModelDb("Jian_2025_peginterferon_alfa_2b")()$population`.

# Source trace

Per-parameter provenance is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Jian_2025_peginterferon_alfa_2b.R`. The
table below collects the whole final model in one place. All values come from
the **"Final model"** block of Jian 2025 Table 2 unless noted.

The paper's Supporting Information is on disk and is load-bearing here:
**Data S2** (`PSP4-14-2014-s002.docx`) is the complete final-model **NONMEM
control stream**, and **Data S1** (`PSP4-14-2014-s001.docx`) holds the
covariate-screening Table S1 and Figures S1-S9. The control stream settles
three things the main text leaves ambiguous or misprints: the sign of
Equation 11, the functional form of the CrCL effect, and the scale of the
reported variability percentages. Each is flagged below and detailed under
"Assumptions and deviations".

| Equation / parameter | Value | Source location |
|---|---|---|
| `d/dt(depot)` | n/a | Eq. 5: `dXa/dt = -ka * Xa`, `Xa(0) = Dose` |
| `d/dt(central)` | n/a | Eq. 6 with Eq. 8 (`Drugfree = Cfree * V`) substituted |
| `cfree` (QE quadratic) | n/a | Eq. 7 |
| `bound` | n/a | Eq. 7 mass balance (`Ctol - Cfree`), identical to `Rtol*Cfree/(KD+Cfree)` under quasi-equilibrium |
| `d/dt(total_target)` | n/a | Eq. 9, with `Rtol(0) = R0` |
| `ksyn` | n/a | Eqs. 10 and 11; **sign taken from Data S2** `KSYN = REC0*KDEG*EXP(-THETA(9)*TIME)` (Eq. 11 is misprinted without the minus) |
| `lcl` (CL/F) | 0.136 L/h | Table 2, RSE 5.3%, bootstrap 95% CI 0.124-0.148 |
| `lvc` (V/F) | 2.95 L | Table 2, RSE 1.2%, bootstrap 95% CI 2.90-2.96 |
| `lka` (ka, healthy) | 0.0101 1/h | Table 2, RSE 10.7%, bootstrap 95% CI 0.0082-0.0119 |
| `lrbase` (R0) | 0.328 ug/L | Table 2, RSE 25.3%, bootstrap 95% CI 0.255-0.474 |
| `lkint` (kint) | 0.0827 1/h | Table 2, RSE 10.5%, bootstrap 95% CI 0.0706-0.0999 |
| `lkdeg` (kdeg) | 0.544 1/h **FIXED** | Table 2 + footnote (a): fixed to the Besremi (ropeginterferon alfa-2b) popPK result, reference [1] |
| `lkd` (KD) | 0.0493 ug/L **FIXED** | Table 2 + footnote (b): fixed from the healthy-subject model |
| `lkdes` (kR) | 0.0068 1/day | Table 2, RSE 19.9%, bootstrap 95% CI 0.0054-0.0095 |
| `e_crcl_cl` | 0.00504 per mL/min | Table 2 value; **linear** form `(1 + theta*(CrCL-110.38))` from Data S2 `CLB_CLCR` (the Table 2 row label misprints it as `exp()`); reference 110.38 mL/min also named in the Figure 4 caption |
| `e_chb_ka` | -0.405 | Table 2 row `theta CHB on ka [(1 + theta) x ka for CHB patients]`; Section 3.4 "CHB patients had a 40.5% lower ka"; Data S2 `KAPHASE` block |
| `etalcl` | 0.118 (reported 34.4%) | Data S2 `$OMEGA` 1; `100*sqrt(0.118)` = Table 2 IIV CL/F |
| `etalvc` | 0.232 (reported 48.2%) | Data S2 `$OMEGA` 2; `100*sqrt(0.232)` = Table 2 IIV V/F |
| `etalka` | 0.197 (reported 44.4%) | Data S2 `$OMEGA` 3; `100*sqrt(0.197)` = Table 2 IIV ka |
| `etalkint` | 0.272 (reported 52.2%) | Data S2 `$OMEGA` 5; `100*sqrt(0.272)` = Table 2 IIV kint |
| `propSd` | 0.19235 (reported 19.2%) | Data S2 `$SIGMA` 2 = 0.037 variance; `sqrt(0.037)` |
| `addSd` | 0.12288 ug/L (reported 123 pg/mL) | Data S2 `$SIGMA` 1 = 15100 pg^2/mL^2; `sqrt(15100)` = 122.88 pg/mL |

# Structural identities (typical-value ladder)

These checks run on the typical-value model (`rxode2::zeroRe()`) and confirm the
implementation reproduces the covariate algebra and the receptor-decay
arithmetic that Jian 2025 states in prose.

```{r identities}
mod  <- readModelDb("Jian_2025_peginterferon_alfa_2b")
modz <- rxode2::zeroRe(mod)

theta <- ui$theta
ka_healthy <- exp(unname(theta[["lka"]]))
e_chb      <- unname(theta[["e_chb_ka"]])
e_crcl     <- unname(theta[["e_crcl_cl"]])
kdes       <- exp(unname(theta[["lkdes"]]))

# (1) CHB absorption is 40.5% slower than healthy (Section 3.4).
ka_chb <- ka_healthy * (1 + e_chb)
stopifnot(isTRUE(all.equal(ka_chb / ka_healthy, 0.595, tolerance = 1e-12)))

# (2) Receptor-synthesis decay: Jian 2025 Discussion states a half-decay of
#     14.5 weeks and ~10% of baseline remaining after 48 weeks of dosing.
half_decay_weeks <- log(2) / kdes / 7
frac_at_48wk     <- exp(-kdes * 48 * 7)
stopifnot(abs(half_decay_weeks - 14.5) < 0.1, abs(frac_at_48wk - 0.10) < 0.01)

# (3) CrCL forest-plot ratios (Section 3.5): 0.904 and 1.21 relative to the
#     110.38 mL/min reference. The effect is LINEAR in the Data S2 control
#     stream, so back-solving the published ratios pins the 10th and 90th CrCL
#     percentiles of the modelling dataset.
crcl_p10 <- (0.904 - 1) / e_crcl + 110.38
crcl_p90 <- (1.21  - 1) / e_crcl + 110.38
stopifnot(
  abs((1 + e_crcl * (crcl_p10 - 110.38)) - 0.904) < 1e-6,
  abs((1 + e_crcl * (crcl_p90 - 110.38)) - 1.21)  < 1e-6
)

tibble::tibble(
  Check = c("ka(CHB) / ka(healthy)",
            "Receptor-synthesis half-decay (weeks)",
            "Receptor synthesis remaining at week 48",
            "CrCL 10th-percentile CL ratio",
            "CrCL 90th-percentile CL ratio"),
  Model = c(sprintf("%.3f", ka_chb / ka_healthy),
            sprintf("%.2f", half_decay_weeks),
            sprintf("%.3f", frac_at_48wk),
            sprintf("%.3f (at CrCL %.1f mL/min)", 0.904, crcl_p10),
            sprintf("%.3f (at CrCL %.1f mL/min)", 1.21,  crcl_p90)),
  Published = c("0.595 (40.5% lower, Section 3.4)",
                "14.5 (Discussion)",
                "~0.10 (Discussion)",
                "0.904 (95% CI 0.842-0.967, Section 3.5)",
                "1.21 (95% CI 1.07-1.34, Section 3.5)")
) |>
  knitr::kable(caption = "Typical-value structural identities against Jian 2025.")
```

The back-solved 10th / 90th CrCL percentiles (`r sprintf("%.1f", crcl_p10)` and
`r sprintf("%.1f", crcl_p90)` mL/min) sit comfortably inside the observed CrCL
range of the modelling dataset (73.2-187.0 mL/min, Jian 2025 Table 1), which is
an independent consistency check on the centering constant.

# Virtual cohorts

Original observed data are not publicly available. Following Jian 2025
Section 2.4, virtual subjects are generated from the estimated IIV distribution
and assigned **fixed** covariate values per arm (the paper assigns "median,
lower, and higher quartiles of covariate ... to virtual patients"). All cohorts
are 100-120 subjects per arm.

```{r cohort}
set.seed(20251202)

N_SINGLE <- 100L   # per single-dose arm
N_MULTI  <- 120L   # per multiple-dose arm
CRCL_REF <- 110.38

# Observation grid: dense through absorption and peak, coarser in the tail.
grid_single <- unique(c(seq(0, 72, by = 2), seq(76, 240, by = 4), seq(248, 720, by = 8)))

make_single <- function(n, dose, healthy, crcl, arm, id_offset) {
  subj <- tibble::tibble(
    id = id_offset + seq_len(n),
    CRCL = crcl, DIS_HEALTHY = healthy, dose_ug = dose, arm = arm
  )
  dosing <- subj |> mutate(time = 0, amt = dose, evid = 1L, cmt = "depot")
  obs    <- subj |> tidyr::crossing(time = grid_single) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
  bind_rows(dosing, obs) |> arrange(id, time, desc(evid))
}

single_arms <- tibble::tribble(
  ~dose, ~healthy, ~crcl,     ~arm,
  45,    1L,       CRCL_REF,  "Healthy 45 ug",
  90,    1L,       CRCL_REF,  "Healthy 90 ug",
  135,   1L,       CRCL_REF,  "Healthy 135 ug",
  180,   1L,       CRCL_REF,  "Healthy 180 ug",
  270,   1L,       CRCL_REF,  "Healthy 270 ug",
  180,   0L,       CRCL_REF,  "CHB 180 ug"
)

ev_single <- do.call(bind_rows, lapply(seq_len(nrow(single_arms)), function(i) {
  a <- single_arms[i, ]
  make_single(N_SINGLE, a$dose, a$healthy, a$crcl, a$arm, (i - 1L) * N_SINGLE)
}))
stopifnot(!anyDuplicated(ev_single[, c("id", "time", "evid")]))

# Typical-value companion: `zeroRe()` makes every subject in an arm identical,
# so one subject per arm carries all the information. Keeping it to one also
# means every downstream summary is one row per arm, which lets the assertions
# below index by arm rather than by row position.
ev_single_typ <- do.call(bind_rows, lapply(seq_len(nrow(single_arms)), function(i) {
  a <- single_arms[i, ]
  make_single(1L, a$dose, a$healthy, a$crcl, a$arm, i)
}))
```

For the 48-week regimen only weeks 1, 8 and 48 are observed -- the three weeks
Jian 2025 Figure 1c reports -- which keeps the output small without changing the
integration.

```{r cohort-multi}
WEEKS_OBS <- c(1L, 8L, 48L)
grid_multi <- unlist(lapply(WEEKS_OBS, function(w) seq((w - 1) * 168, w * 168, by = 4)))
grid_multi <- sort(unique(grid_multi))

make_multi <- function(n, healthy, crcl, arm, id_offset) {
  subj <- tibble::tibble(
    id = id_offset + seq_len(n),
    CRCL = crcl, DIS_HEALTHY = healthy, dose_ug = 180, arm = arm
  )
  dosing <- subj |> tidyr::crossing(time = seq(0, 47 * 168, by = 168)) |>
    mutate(amt = 180, evid = 1L, cmt = "depot")
  obs <- subj |> tidyr::crossing(time = grid_multi) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
  bind_rows(dosing, obs) |> arrange(id, time, desc(evid))
}

multi_arms <- tibble::tribble(
  ~healthy, ~crcl,     ~arm,
  1L,       CRCL_REF,  "Healthy, CrCL 110",
  0L,       CRCL_REF,  "CHB, CrCL 110",
  0L,       crcl_p10,  "CHB, CrCL 91 (10th pctile)",
  0L,       crcl_p90,  "CHB, CrCL 152 (90th pctile)"
)

ev_multi <- do.call(bind_rows, lapply(seq_len(nrow(multi_arms)), function(i) {
  a <- multi_arms[i, ]
  make_multi(N_MULTI, a$healthy, a$crcl, a$arm, 10000L + (i - 1L) * N_MULTI)
}))
stopifnot(!anyDuplicated(ev_multi[, c("id", "time", "evid")]))

ev_multi_typ <- do.call(bind_rows, lapply(seq_len(nrow(multi_arms)), function(i) {
  a <- multi_arms[i, ]
  make_multi(1L, a$healthy, a$crcl, a$arm, i)
}))
```

# Simulation

```{r simulate}
KEEP <- c("arm", "CRCL", "DIS_HEALTHY", "dose_ug")

sim_single <- rxode2::rxSolve(mod, events = as.data.frame(ev_single), keep = KEEP) |>
  as.data.frame()
sim_multi  <- rxode2::rxSolve(mod, events = as.data.frame(ev_multi),  keep = KEEP) |>
  as.data.frame()

# Typical-value (no IIV) companions for the deterministic panels.
simz_single <- rxode2::rxSolve(modz, events = as.data.frame(ev_single_typ), keep = KEEP) |>
  as.data.frame()
simz_multi  <- rxode2::rxSolve(modz, events = as.data.frame(ev_multi_typ),  keep = KEEP) |>
  as.data.frame()

# One typical subject per arm, so every per-arm summary below is a single row.
stopifnot(
  simz_single |> count(arm) |> pull(n) |> (\(x) length(unique(x)) == 1L)(),
  n_distinct(simz_single$id) == nrow(single_arms),
  n_distinct(simz_multi$id)  == nrow(multi_arms)
)

stopifnot(all(is.finite(sim_single$Cc[!is.na(sim_single$Cc)])),
          all(is.finite(sim_multi$Cc[!is.na(sim_multi$Cc)])),
          all(sim_single$Cc[!is.na(sim_single$Cc)] >= 0))
```

```{r auc-helper}
# Linear-trapezoid AUC over an explicit window.
auc_window <- function(time, conc, from, to) {
  keep <- time >= from & time <= to
  tt <- time[keep]; cc <- conc[keep]
  o <- order(tt); tt <- tt[o]; cc <- cc[o]
  sum(diff(tt) * (head(cc, -1) + tail(cc, -1)) / 2)
}
```

# Replicate published figures

## Figure 1d, 1e -- single-dose non-proportionality

Jian 2025 Figure 1d,e show AUC/dose rising and CL/F falling with dose, the
exploratory signal that motivated a saturable (TMDD) elimination component.

```{r figure-1de}
np <- simz_single |>
  filter(!is.na(Cc), grepl("^Healthy", arm)) |>
  group_by(arm, dose_ug) |>
  summarise(auc = auc_window(time, Cc, 0, 720), .groups = "drop") |>
  # group_by() returns rows sorted by the grouping key as a STRING, which puts
  # "Healthy 135 ug" before "Healthy 45 ug"; sort numerically before any
  # position-dependent assertion below.
  arrange(dose_ug) |>
  mutate(auc_per_dose = auc / dose_ug / 24 * 1000,   # (pg/mL)*day per ug
         clf_per_day  = dose_ug / auc * 24)          # L/day

np |>
  select(dose_ug, auc_per_dose, clf_per_day) |>
  tidyr::pivot_longer(-dose_ug) |>
  mutate(name = recode(name,
                       auc_per_dose = "AUC/dose ((pg/mL)*day per ug)",
                       clf_per_day  = "CL/F (L/day)")) |>
  ggplot(aes(dose_ug, value)) +
  geom_point(size = 2.5) + geom_line() +
  facet_wrap(~name, scales = "free_y") +
  labs(x = "Dose (ug)", y = NULL,
       title = "Figure 1d,e -- dose non-proportionality after a single dose",
       caption = "Replicates Figure 1d,e of Jian 2025 (typical-value healthy subject, CrCL 110.38 mL/min).")

stopifnot(
  all(diff(np$auc_per_dose) > 0),  # AUC/dose increases with dose (Fig 1d)
  all(diff(np$clf_per_day)  < 0)   # CL/F decreases with dose (Fig 1e)
)
knitr::kable(np |> select(`Dose (ug)` = dose_ug,
                          `AUC/dose ((pg/mL)*day/ug)` = auc_per_dose,
                          `CL/F (L/day)` = clf_per_day),
             digits = 1,
             caption = "Saturable elimination: AUC/dose rises and CL/F falls across the studied dose range.")
```

## Figure 1c -- exposure rises from week 1 to week 8 to week 48

The receptor-downregulation mechanism was added precisely because observed
week-48 exposure exceeded week-8 exposure, which a classical TMDD model cannot
produce (Jian 2025 Section 3.2 and Figure 1c).

```{r figure-1c}
interval_auc <- function(df) {
  df |>
    filter(!is.na(Cc)) |>
    group_by(arm, id) |>
    reframe(week = WEEKS_OBS,
            auctau = vapply(WEEKS_OBS, function(w)
              auc_window(time, Cc, (w - 1) * 168, w * 168), numeric(1))) |>
    mutate(auc_per_dose = auctau / 180 / 24 * 1000)
}

wk <- interval_auc(sim_multi)

wk |>
  filter(arm %in% c("Healthy, CrCL 110", "CHB, CrCL 110")) |>
  mutate(week = factor(week, levels = WEEKS_OBS, labels = paste("Week", WEEKS_OBS))) |>
  ggplot(aes(week, auc_per_dose, fill = arm)) +
  geom_boxplot(outlier.size = 0.4, alpha = 0.75) +
  labs(x = NULL, y = "AUCtau/dose ((pg/mL)*day per ug)", fill = NULL,
       title = "Figure 1c -- dosing-interval exposure at weeks 1, 8 and 48",
       caption = "Replicates Figure 1c of Jian 2025 (180 ug weekly).")

wk_typ <- interval_auc(simz_multi) |>
  filter(arm %in% c("Healthy, CrCL 110", "CHB, CrCL 110")) |>
  select(arm, week, auctau, auc_per_dose)

# Paper's claim: exposure increases monotonically week 1 -> 8 -> 48.
stopifnot(
  wk_typ |> group_by(arm) |> arrange(week, .by_group = TRUE) |>
    summarise(ok = all(diff(auctau) > 0), .groups = "drop") |> pull(ok) |> all()
)
knitr::kable(wk_typ |> tidyr::pivot_wider(id_cols = arm, names_from = week,
                                          values_from = auctau,
                                          names_prefix = "Week ") |>
               rename("Arm" = arm),
             digits = 0,
             caption = "Typical-value AUCtau (ug*h/L) per dosing interval. Exposure rises as receptor synthesis decays.")
```

## Figure 5e-h -- health status at first dose and at steady state

Jian 2025 reports that the slower CHB absorption lowers both Cmax and AUC after
the **first** dose (Figure 5e,f) but that, by steady state, receptor
downregulation has made elimination essentially linear so the **steady-state
AUC is comparable** between the two populations, with healthy subjects retaining
a higher Cmax and larger peak-to-trough fluctuation (Figure 5g,h).

```{r figure-5gh}
hs <- c("Healthy, CrCL 110", "CHB, CrCL 110")

simz_multi |>
  filter(!is.na(Cc), arm %in% hs, time >= 47 * 168) |>
  mutate(time_d = (time - 47 * 168) / 24) |>
  ggplot(aes(time_d, Cc, colour = arm)) +
  geom_line(linewidth = 0.9) +
  labs(x = "Time within the week-48 dosing interval (days)", y = "Cc (ug/L)", colour = NULL,
       title = "Figure 5g -- steady-state (week 48) concentration-time profile",
       caption = "Replicates Figure 5g of Jian 2025 (180 ug weekly, typical-value subject).")

ss <- simz_multi |>
  filter(!is.na(Cc), arm %in% hs, time >= 47 * 168) |>
  group_by(arm) |>
  summarise(auctau = auc_window(time, Cc, 47 * 168, 48 * 168),
            cmax = max(Cc), ctrough = min(Cc), .groups = "drop") |>
  mutate(fluctuation = cmax / ctrough)

fd <- simz_single |>
  filter(!is.na(Cc), arm %in% c("Healthy 180 ug", "CHB 180 ug")) |>
  group_by(arm) |>
  summarise(auc = auc_window(time, Cc, 0, 720), cmax = max(Cc), .groups = "drop")

cl_typ  <- exp(unname(theta[["lcl"]]))
dose_cl <- 180 / cl_typ

# (a) First dose: healthy > CHB on BOTH Cmax and AUC (Figure 5e,f).
stopifnot(
  fd$cmax[fd$arm == "Healthy 180 ug"] > fd$cmax[fd$arm == "CHB 180 ug"],
  fd$auc [fd$arm == "Healthy 180 ug"] > fd$auc [fd$arm == "CHB 180 ug"]
)
# (b) Steady state: AUCtau comparable between populations (within 1%).
stopifnot(abs(diff(ss$auctau)) / mean(ss$auctau) < 0.01)
# (c) Steady state: elimination is essentially linear, so AUCtau ~ Dose/CL.
stopifnot(all(abs(ss$auctau / dose_cl - 1) < 0.05))
# (d) Healthy retains the higher Cmax and the larger fluctuation.
stopifnot(
  ss$cmax[ss$arm == "Healthy, CrCL 110"] > ss$cmax[ss$arm == "CHB, CrCL 110"],
  ss$fluctuation[ss$arm == "Healthy, CrCL 110"] > ss$fluctuation[ss$arm == "CHB, CrCL 110"]
)

bind_rows(
  fd |> transmute(Phase = "First dose", Arm = arm, `AUC (ug*h/L)` = auc,
                  `Cmax (ug/L)` = cmax, `Peak/trough` = NA_real_),
  ss |> transmute(Phase = "Steady state (week 48)", Arm = arm,
                  `AUC (ug*h/L)` = auctau, `Cmax (ug/L)` = cmax,
                  `Peak/trough` = fluctuation)
) |>
  knitr::kable(digits = 2,
               caption = sprintf(paste("Health status at first dose vs steady state (typical value, 180 ug).",
                                       "Steady-state Dose/CL = %.0f ug*h/L."), dose_cl))
```

The steady-state AUC values are identical to three significant figures because
health status acts only on `ka`, and by week 48 the residual receptor pool is
too small for the saturable route to matter -- exactly the mechanism Jian 2025
invokes in the Discussion ("the total elimination of Pegbing at steady state
primarily relied on linear clearance ... comparable steady-state AUCs of Pegbing
were estimated between CHB patients and healthy subjects").

```{r tmdd-fraction}
# Quantify that claim: fraction of total elimination flowing through the
# saturable (TMDD) route, at week 1 vs week 48, for the typical CHB subject.
kint_t <- exp(unname(theta[["lkint"]]))
vc_t   <- exp(unname(theta[["lvc"]]))

frac_tmdd <- simz_multi |>
  filter(!is.na(Cc), arm == "CHB, CrCL 110") |>
  mutate(week = cut(time, breaks = c(-1, 168, 8 * 168, 48 * 168),
                    labels = c("Week 1", "Week 8", "Week 48")),
         bound = Cc - cfree,
         rate_tmdd = kint_t * bound * vc_t,
         rate_lin  = cl_typ * cfree) |>
  filter(!is.na(week)) |>
  group_by(week) |>
  summarise(`TMDD share of elimination` = sum(rate_tmdd) / sum(rate_tmdd + rate_lin),
            .groups = "drop")

stopifnot(
  frac_tmdd$`TMDD share of elimination`[frac_tmdd$week == "Week 1"] >
    frac_tmdd$`TMDD share of elimination`[frac_tmdd$week == "Week 48"]
)
knitr::kable(frac_tmdd, digits = 3,
             caption = "The saturable elimination route fades as the receptor pool is downregulated.")
```

## Figure 5a-d -- creatinine clearance

Jian 2025 reports that CrCL differences are negligible after the first dose but
become visible at steady state (Figure 5a-d), and that the effect nevertheless
fails the 80-120% clinical-relevance band.

```{r figure-5cd}
crcl_arms <- c("CHB, CrCL 91 (10th pctile)", "CHB, CrCL 110", "CHB, CrCL 152 (90th pctile)")

sim_multi |>
  filter(!is.na(Cc), arm %in% crcl_arms, time >= 47 * 168) |>
  mutate(time_d = (time - 47 * 168) / 24) |>
  group_by(arm, time_d) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
            .groups = "drop") |>
  ggplot(aes(time_d, Q50, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.9) +
  labs(x = "Time within the week-48 dosing interval (days)", y = "Cc (ug/L)",
       colour = NULL, fill = NULL,
       title = "Figure 5c -- steady-state exposure by creatinine clearance",
       caption = "Replicates Figure 5c of Jian 2025 (CHB, 180 ug weekly; median and 90% prediction interval).")

crcl_ss <- simz_multi |>
  filter(!is.na(Cc), arm %in% crcl_arms, time >= 47 * 168) |>
  group_by(arm, CRCL) |>
  summarise(auctau = auc_window(time, Cc, 47 * 168, 48 * 168), .groups = "drop") |>
  arrange(CRCL) |>
  mutate(ratio_vs_ref = auctau / auctau[which.min(abs(CRCL - CRCL_REF))])

# Higher CrCL -> higher clearance -> lower steady-state exposure.
stopifnot(all(diff(crcl_ss$auctau) < 0))
# At steady state elimination is linear, so the AUC ratio should be the
# reciprocal of the published CL ratio.
published_cl_ratio <- c(0.904, 1.0, 1.21)
stopifnot(max(abs(crcl_ss$ratio_vs_ref - 1 / published_cl_ratio)) < 0.02)

knitr::kable(crcl_ss |> rename("Arm" = arm, "CrCL (mL/min)" = CRCL,
                               "AUCtau (ug*h/L)" = auctau,
                               "AUC ratio vs reference" = ratio_vs_ref),
             digits = 3,
             caption = "Steady-state exposure vs CrCL; the AUC ratio is the reciprocal of the published CL ratio (0.904 / 1.21).")
```

# PKNCA validation

NCA is run on the single-dose arms. Following the paper's bioanalytical method,
each subject's profile is truncated at the last point at or above the ELISA
LLOQ of 300 pg/mL (= 0.3 ug/L); this both matches what the assay could see and
keeps the terminal-slope fit out of the solver-noise tail.

```{r pknca}
LLOQ <- 0.3

sim_nca <- sim_single |>
  filter(!is.na(Cc)) |>
  group_by(id) |>
  filter(time <= max(time[Cc >= LLOQ])) |>
  ungroup() |>
  select(id, time, Cc, arm, dose_ug)

# Guarantee a time-zero row per subject (extravascular pre-dose Cc = 0).
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, arm, dose_ug) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, arm, time, .keep_all = TRUE) |>
  arrange(id, time)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(sim_nca), Cc ~ time | arm + id)

dose_df <- ev_single |>
  filter(evid == 1) |>
  select(id, time, amt, arm)
dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | arm + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE,
  half.life = TRUE, cl.obs = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_summary <- as.data.frame(nca_res) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life", "cl.obs")) |>
  group_by(arm, PPTESTCD) |>
  summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median)

nca_summary |>
  mutate(half.life_d = half.life / 24) |>
  rename("Arm" = arm, "Cmax (ug/L)" = cmax, "Tmax (h)" = tmax,
         "AUCinf (ug*h/L)" = aucinf.obs, "t1/2 (h)" = half.life,
         "t1/2 (days)" = half.life_d, "CL/F (L/h)" = cl.obs) |>
  knitr::kable(digits = 2,
               caption = "Median simulated single-dose NCA by arm (100 virtual subjects per arm).")
```

## Comparison against published NCA

Jian 2025 does not publish an NCA table; the only NCA-comparable number in the
paper is the post-hoc elimination half-life quoted in the Discussion ("the
elimination half-life of Pegbing is about 7 days"). It is compared below.

```{r nca_compare}
published <- tibble::tribble(
  ~arm,              ~half.life,
  "Healthy 180 ug",  7 * 24,
  "CHB 180 ug",      7 * 24
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "arm",
  params        = "half.life",
  units         = c(half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(cmp,
             caption = "Simulated vs published terminal half-life. * differs from the reference by >20%.")
```

**This row is starred, and the discrepancy is real rather than an artefact of
the simulation.** With `ka` (0.0101 1/h healthy, 0.00601 1/h CHB) far smaller
than `kel = CL/V` (0.0461 1/h), the model is firmly flip-flop: the terminal
slope is set by absorption, giving `log(2)/ka` = 68.6 h (2.9 days) in healthy
subjects and 115 h (4.8 days) in CHB patients. TMDD only steepens the true
terminal phase further. No combination of the published Table 2 estimates
yields a 7-day terminal half-life; see "Assumptions and deviations" below. No
parameter has been adjusted to close the gap.

```{r flipflop-check}
ka_h <- ka_healthy; ka_c <- ka_chb
kel  <- cl_typ / exp(unname(theta[["lvc"]]))
stopifnot(ka_h < kel, ka_c < kel)   # flip-flop confirmed
tibble::tibble(
  Quantity = c("ka healthy (1/h)", "ka CHB (1/h)", "kel = CL/V (1/h)",
               "log(2)/ka healthy (days)", "log(2)/ka CHB (days)",
               "log(2)/kel (days)"),
  Value = c(ka_h, ka_c, kel, log(2) / ka_h / 24, log(2) / ka_c / 24, log(2) / kel / 24)
) |>
  knitr::kable(digits = 4,
               caption = "Absorption is rate-limiting, so the terminal slope reflects ka, not CL/V.")
```

# Assumptions and deviations

- **Equation 11 sign (corrected from the control stream).** As typeset,
  Jian 2025 Eq. 11 reads `ksyn = kdeg * R0 * exp(kR * t)` with **no minus
  sign** (confirmed against the published PDF; minus signs render correctly in
  Eqs. 5, 6 and 9 of the same paper). The Data S2 final-model control stream
  settles it directly -- it codes
  `KSYN = REC0*KDEG*EXP(-THETA(9)*TIME)` -- so the executed model decays and
  the typeset equation is simply missing its minus. The model implements the
  decay form. The same conclusion follows four more ways from the paper's own
  content, independent of the supplement:
  (i) the sentence introducing Eq. 11 -- "the synthesis rate of receptors
  **decreases** exponentially from the beginning of treatment"; (ii) the
  Figure 2 caption -- "kR, exponential **decrease** rate of IFN receptor";
  (iii) the Discussion's "decrease by half after 14.5 weeks", which reproduces
  `log(2)/0.0068 per day = 14.56 weeks` only under decay; and (iv) the
  Discussion's "reduced to approximately 10% of baseline" after 48 weeks, which
  reproduces `exp(-0.0068 * 336) = 0.102` only under decay. With the printed
  `+` sign the receptor pool would grow to roughly 9.8x baseline by week 48,
  inverting the very observation (Figure 1c) the mechanism was introduced to
  explain.
- **Published terminal half-life is not reproducible from Table 2.** The
  Discussion states "the elimination half-life of Pegbing is about 7 days";
  the Table 2 parameter set implies 2.9 days (healthy) to 4.8 days (CHB), as
  shown above. The parameters are used as published.
- **Table 2 back-transform footnote (settled by `$OMEGA` / `$SIGMA`).** The
  abbreviations footnote defines the reported percentages as
  `SQRT(omega^2 - 1)` and `SQRT(sigma^2 - 1)`, which is impossible for these
  estimates (it would need `omega^2 > 1`). Two readings were plausible from
  the main text alone: the log-normal CV `SQRT(exp(omega^2) - 1)`, or the
  log-scale SD `SQRT(omega^2)`. The Data S2 `$OMEGA` block decides it -- and
  the two readings differ by up to 7 percentage points, so this is not a
  cosmetic choice:

    | Parameter | `$OMEGA` variance | `100*sqrt(var)` | `100*sqrt(exp(var)-1)` | Table 2 |
    |---|---|---|---|---|
    | CL/F | 0.118 | **34.4** | 35.4 | 34.4 |
    | V/F | 0.232 | **48.2** | 51.1 | 48.2 |
    | ka | 0.197 | **44.4** | 46.7 | 44.4 |
    | kint | 0.272 | **52.2** | 55.9 | 52.2 |

  `100*sqrt(variance)` reproduces every printed digit; the log-normal CV
  matches none. The `$SIGMA` variances (0.037 and 15100 pg^2/mL^2) back out to
  19.2% and 123 pg/mL on the same scale. The variances are therefore taken
  from `$OMEGA` / `$SIGMA` verbatim, not back-transformed. The footnote's
  label "CI, confidence interval" attached to that formula is a further slip
  for "CV".
- **CrCL effect is linear, not exponential.** The Jian 2025 Table 2 row label
  prints `CL x exp(theta x (CrCL-110.38))`, but the executed Data S2 control
  stream codes the **linear** relation
  `CLB_CLCR=(1+THETA(11)*(B_CLCR-110.38))` -- Equation 1 of Methods
  Section 2.3, the PsN stepwise-covariate-model linear form. The model follows
  the control stream. Over the observed CrCL range (67.6-187.0 mL/min,
  Table 1) the two forms differ by up to about 6% in `CL`, so the choice is
  visible but not dramatic. The control stream also shows the covariate is
  **baseline** CrCL (`B_CLCR`), not the time-varying `CLCR` column that the
  dataset also carries.
- **Covariate screening path.** Table S1 (Data S1) records the full stepwise
  path: forward inclusion added health status on `ka` (dOFV -15.84), CrCL on
  `CL` (-7.43) and body weight on `V` (-5.77); backward elimination then
  dropped weight on `V` (+5.74, below the 6.63 threshold) while retaining CrCL
  on `CL` (+10.01) and health status on `ka` (+15.77). This confirms the
  two-covariate final model implemented here.
- **Table 2 bootstrap-median typo for R0.** The final-model bootstrap median
  for R0 prints as `0.0328`, ten-fold below the 0.328 point estimate and
  outside its own bootstrap 95% CI (0.255-0.474). The point estimate is used.
- **`ka` units in the Discussion.** The Discussion compares "a higher
  absorption rate constant (0.010/day v.s. 0.00583/day)", but Table 2 labels
  `ka` as `/h` in both the base and final model blocks (and reserves `1/day`
  for `kR`, showing the units were tracked per row). The `/h` reading is used:
  at `0.0101/day` the absorption half-life would be 69 days, which is
  incompatible with the 17-day Phase I sampling window and with the observed
  Tmax.
- **Bioavailability.** Eq. 5 sets `Xa(0) = Dose` with no separate `F` term, so
  `F = 1` and both CL/F and V/F are apparent. Figure 2 draws an `F` but it is
  not identifiable and is not estimated.
- **Observed analyte.** The ELISA quantifies total drug, and Jian 2025
  integrates `Drugtol`; the model therefore reports `Cc = Ctol`. Free
  concentration is available as the derived variable `cfree`. The paper does
  not state the total-vs-free assay specificity explicitly.
- **Receptor state units.** `total_target` carries a drug-equivalent
  *concentration* (ug/L), not an amount: Eq. 9 is written on the concentration
  scale and Table 2 reports R0 in ug/L, even though the symbol glossary
  loosely calls `Rtol` an "amount".
- **`kint < kdeg` stabilises the receptor pool transiently.** With the
  published `kint = 0.0827/h` and the Besremi-fixed `kdeg = 0.544/h`, the
  `-(kint - kdeg) * bound` term of Eq. 9 is *positive* while drug is bound, so
  total receptor rises above R0 early in treatment before the `ksyn` decay
  takes over (it falls to ~63% of R0 by week 48 in the typical subject, while
  the *synthesis rate* falls to ~10%). This is the standard
  Gibiansky QSS/QE `Rtot` equation evaluated at the published constants, not a
  transcription choice; the Discussion's "reduced to approximately 10% of
  baseline" describes the synthesis rate / drug-free equivalent baseline.
- **Covariate polarity.** The canonical register column is `DIS_HEALTHY`
  (1 = healthy), while Jian 2025 uses healthy as the reference and reports the
  effect on the CHB arm. The model evaluates the complement `(1 - DIS_HEALTHY)`
  so the published `(1 + theta)` form is reproduced exactly.
- **Base model excluded.** Jian 2025 Table 2 also reports a base (no-covariate)
  model. Per the library's replicate-the-authors policy, only the final model
  is packaged; the base-model estimates are recorded in the source-trace notes
  only.
- **Body weight is not in the final model.** The Abstract claims "body weight
  affected the volume of distribution", but the Discussion states weight was
  dropped in backward elimination and Table 2's final model has no weight term.
  Weight is recorded in `covariatesDataExcluded`, not `covariateData`. The
  Data S2 control stream corroborates this in a third way: it still contains the
  vestigial line `TVV = VCOV*TVV` from the PsN stepwise-covariate-model template,
  but `VCOV` is never assigned anywhere in `$PK` -- the weight-on-`V` block was
  deleted during backward elimination and only its consumer was left behind. The
  reported `V/F` equals `THETA(2)` exactly (2.95 L), so the executed model
  carried no covariate on `V`. This model therefore applies none.
- **Table 1 validation-cohort sex percentages are transposed** (printed as
  Male 84 (27.0%) / Female 31 (73.0%), but 84/115 = 73.0%). This affects only
  the held-out external-validation cohort description, not the model.
- **Supplement.** Data S1-S3 were retrieved from the Europe PMC
  `supplementaryFiles` endpoint for PMC12706394 and are the source of the
  three corrections above. Data S3 (`PSP4-14-2014-s003.csv`) is a
  single-subject example dataset, not the analysis dataset, so no
  subject-level distribution could be recovered from it; the virtual cohorts
  below use the Table 1 summaries instead.
- **Virtual cohorts.** Covariates are fixed per arm at the values Jian 2025
  uses for its own covariate simulations (Section 2.4), with between-subject
  spread supplied by the estimated IIV. Cohorts are 100 (single-dose) or 120
  (multiple-dose) subjects per arm.
