---
title: "Serotonergic thermoregulation in mice: harmaline + 5-MeO-DMT (Jiang 2016)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Serotonergic thermoregulation in mice: harmaline + 5-MeO-DMT (Jiang 2016)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

```{r libraries, message = FALSE}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(ggplot2)
```

## The model

Jiang XL, Shen HW, Mager DE, Schmidt S, Yu AM. *Development of a
mechanism-based pharmacokinetic/pharmacodynamic model to characterize the
thermoregulatory effects of serotonergic drugs in mice.* Acta Pharm Sin B.
2016;6(5):492-503. [doi:10.1016/j.apsb.2016.07.007](https://doi.org/10.1016/j.apsb.2016.07.007)

The paper builds one integrated PK/PD model (its Fig. 1) linking the
pharmacokinetics of the monoamine oxidase-A inhibitor **harmaline** and the
5-HT receptor agonist **5-MeO-DMT** to core body temperature (CBT) in mice.
Three things are happening at once:

* a **pharmacokinetic** interaction -- harmaline competitively inhibits two of
  the three routes that clear 5-MeO-DMT, so harmaline raises 5-MeO-DMT exposure;
* a **pharmacodynamic** interaction -- harmaline lowers the 5-MeO-DMT
  concentration producing half-maximal thermogenesis roughly 4-fold;
* a **genotype** effect -- CYP2D6-humanized transgenic mice clear harmaline
  faster than wild-type mice, which blunts harmaline's hypothermia.

```{r load-model}
mod <- rxode2::rxode(readModelDb("Jiang_2016_harmaline_meodmt_mouse"))
mod$state
```

## Population

```{r population}
pop <- mod$population
knitr::kable(
  data.frame(Field = names(pop), Value = unlist(lapply(pop, as.character))),
  row.names = FALSE
)
```

Male FVB/N wild-type and Tg-*CYP2D6* mice, 25-35 g, dosed intraperitoneally.
Core body temperature was recorded by implanted telemetry at 10 samples per
minute and averaged over 5- or 10-minute bins for modelling. The fit was a
naive-pooled maximum-likelihood analysis in ADAPT V, so the model carries
typical values only -- there is no inter-individual variability to simulate and
every check below is deterministic.

## Source trace

Every value in `ini()` comes from the locations below.

```{r source-trace, echo = FALSE}
knitr::kable(
  tibble::tribble(
    ~Quantity, ~Source,
    "Harmaline PK: ka-H, VC-H, VP-H, CLD-H, CLother-H, CLCYP2D6-H, FH", "Methods 2.4, p. 494 (values carried over from the companion PK study, ref. 27, and held fixed here)",
    "5-MeO-DMT PK: ka-M, VC-M, VP-M, CLD-M, FM", "Methods 2.4, p. 494",
    "5-MeO-DMT elimination: Vmax and Km for MAO-A, other murine, O-demethylation", "Methods 2.4, p. 494",
    "fmCYP2D6(D)-M = 30.1%", "Methods 2.4, p. 494; the multiplicative form Vmax(D)-M x (1 + fm) is printed inside the Fig. 1 schematic",
    "Ki(M)-H = 0.048 umol/L", "Methods 2.4, p. 494 (literature value, ref. 32)",
    "Ki(D)-H = 7.13 umol/L", "Methods 2.4, p. 495 (from the upstream PK model fitting)",
    "Competitive (on Km) form of both inhibition terms", "Fig. 1: both Ki arrows terminate in an inhibition bar on the Km label, not the Vmax label",
    "Baseline CBT equation", "Eq. (1)",
    "Indirect-response turnover with adaptive feedback", "Eq. (2), closed at time zero by Eq. (3)",
    "Handling/injection stress signal", "Eq. (4)",
    "Harmaline stimulation of heat loss (5-HT1A)", "Eq. (5)",
    "5-MeO-DMT transduction chain (3 transit compartments)", "Eqs. (6)-(8)",
    "Final thermoregulatory ODE", "Eq. (9)",
    "Residual-error structure", "Eq. (10) gives the form; no sigma value is printed anywhere in the paper",
    "All PD parameter values and their CV%", "Table 1"
  )
)
```

### Units

Every published volume, clearance and Vmax is weight-normalised, so the whole
system is per kg body weight and compartment amounts are umol/kg.

```{r units-table, echo = FALSE}
knitr::kable(
  tibble::tribble(
    ~Term, ~Units, ~Checks,
    "central_harmaline", "umol/kg", "amount per kg",
    "Cc_harmaline = central_harmaline / vc_harmaline", "umol/L", "(umol/kg) / (L/kg)",
    "cl_harmaline * Cc_harmaline", "umol/min/kg", "(L/min/kg) x (umol/L)",
    "elim_meodmt", "umol/min/kg", "Vmax already umol/min/kg; the MM fraction is unitless",
    "sH = ks_harmaline * Cc_harmaline", "unitless", "(L/umol) x (umol/L)",
    "kin = kout * rbase", "degC/min", "(1/min) x degC",
    "kin * (tempBasal / temp)", "degC/min", "feedback ratio is unitless",
    "kout * (1 + sH) * temp", "degC/min", "(1/min) x degC"
  )
)
```

Note the one unit inconsistency in the source: `CLother-H` and `CLCYP2D6-H` are
printed as "L/min" while every other volume and clearance in the same paragraph
carries "/kg". They must be L/min/**kg** for the model to be dimensionally
consistent, and the check under "Harmaline pharmacokinetics" below confirms this
against the paper's own simulated concentrations.

## Setting up simulations

Doses are published in mg/kg; the model works in umol/kg. The two molecular
weights are chemical constants and are not taken from the paper.

```{r helpers}
MW_HARMALINE <- 214.26 # g/mol, harmaline free base
MW_MEODMT <- 218.30 # g/mol, 5-MeO-DMT free base

# One arm of the experiment. `inj_extra` adds a second handling/injection event
# at `t_second` even when nothing pharmacological is given (the vehicle arms).
simulate_arm <- function(label, dose_harmaline = 0, dose_meodmt = 0, tg = 0,
                         t_second = 15, inj_extra = FALSE, tmax = 240,
                         dt = 0.5) {
  inj_times <- numeric(0)
  ev <- rxode2::et(seq(0, tmax, by = dt), cmt = "temp")

  if (dose_harmaline > 0) {
    ev <- rxode2::et(
      ev,
      amt = dose_harmaline / MW_HARMALINE * 1000,
      cmt = "depot_harmaline", time = 0
    )
    inj_times <- c(inj_times, 0)
  }
  if (dose_meodmt > 0) {
    ev <- rxode2::et(
      ev,
      amt = dose_meodmt / MW_MEODMT * 1000,
      cmt = "depot_meodmt", time = t_second
    )
    inj_times <- c(inj_times, t_second)
  }
  if (inj_extra) inj_times <- c(inj_times, t_second)

  # One unit into `stress` per handling event; the model's f(stress) scales it
  # to the published magnitude S0 = 0.265.
  for (tt in sort(unique(inj_times))) {
    ev <- rxode2::et(ev, amt = 1, cmt = "stress", time = tt)
  }

  d <- as.data.frame(ev)
  d$CYP2D6_TG <- tg
  # kS-H switches on the handling burden, not on what was injected.
  d$INJ_REPEAT <- as.integer(length(unique(inj_times)) > 1)
  # SC50-M switches on whether harmaline was actually present.
  d$CONMED_HARMALINE <- as.integer(dose_harmaline > 0 && dose_meodmt > 0)
  d$DOSE_HARMALINE_MGKG <- dose_harmaline

  out <- rxode2::rxSolve(mod, d, returnType = "data.frame")
  out$arm <- label
  out$genotype <- ifelse(tg == 1, "Tg-CYP2D6", "Wild-type")
  out$dose_harmaline <- dose_harmaline
  out$dose_meodmt <- dose_meodmt
  out
}
```

## Check 1: the undosed system holds its baseline

With nothing administered, Eq. (9) collapses to
`kin * (tempBasal / temp) = kout * temp`. Because `kin = kout * rbase` is a
*constant* fixed at time zero by Eq. (3), the quasi-steady-state solution is the
geometric mean `temp(t) = sqrt(rbase * tempBasal(t))` -- not `tempBasal(t)`
itself. This is a purely deterministic identity, so it is asserted tightly.

First the exact case. Override the baseline drift to zero and the attractor
becomes the constant `rbase`, which the system must hold indefinitely. Nothing
here is stochastic, so this is asserted at machine precision.

```{r check-baseline-exact}
rbase <- 35.8
drift <- 0.00322

ev_base <- rxode2::et(seq(0, 300, by = 0.5), cmt = "temp")
d_base <- as.data.frame(ev_base)
d_base$CYP2D6_TG <- 0
d_base$INJ_REPEAT <- 0
d_base$CONMED_HARMALINE <- 0
d_base$DOSE_HARMALINE_MGKG <- 0

flat <- rxode2::rxSolve(mod, d_base, params = c(drift_rbase = 0),
                        returnType = "data.frame")
max_flat_err <- max(abs(flat$temp - rbase))

stopifnot(max_flat_err < 1e-8)
c(max_deviation_from_baseline = max_flat_err)
```

With the published drift restored, `tempBasal` becomes a moving target and the
turnover system tracks it with a small lag, so `sqrt(rbase * tempBasal(t))` is
the quasi-steady state rather than an exact solution. The gap is the lag of a
finite-rate system, and it is bounded.

```{r check-baseline-drift}
base_sim <- simulate_arm("baseline", tmax = 300)
quasi_ss <- sqrt(rbase * (rbase + drift * base_sim$time))
max_lag <- max(abs(base_sim$temp - quasi_ss))

stopifnot(
  # Deterministic: realised lag is 0.0275 degC, bounded here with headroom.
  max_lag < 0.05,
  # The system starts exactly at the published baseline.
  abs(base_sim$temp[1] - rbase) < 1e-8
)
c(max_lag_behind_quasi_ss = max_lag, temp_at_300min = tail(base_sim$temp, 1))
```

The undosed trajectory rises from 35.8 degC to `r round(tail(base_sim$temp, 1), 2)`
degC over 300 min. This matters as a structural check: reading Eq. (9)
with a *time-varying* `kin = kout * tempBasal(t)` would instead put the
300-minute baseline at 36.77 degC. Fig. 2's fitted baseline line ends near
36.3 degC, which is the constant-`kin` reading the model implements.

## Check 2: perturbation recovery

Displacing the temperature state and running forward must bring it back to the
same attractor.

```{r check-recovery}
ev_free <- rxode2::et(seq(0, 300, by = 1), cmt = "temp")
d_free <- as.data.frame(ev_free)
d_free$CYP2D6_TG <- 0
d_free$INJ_REPEAT <- 0
d_free$CONMED_HARMALINE <- 0
d_free$DOSE_HARMALINE_MGKG <- 0

recover <- function(start) {
  s <- rxode2::rxSolve(mod, d_free, inits = c(temp = start),
                       returnType = "data.frame")
  tail(s$temp, 1)
}
ends <- vapply(c(0.8 * rbase, rbase, 1.2 * rbase), recover, numeric(1))
stopifnot(max(abs(ends - tail(base_sim$temp, 1))) < 1e-3)
round(ends, 4)
```

All three start points converge on the same trajectory, so the feedback term has
a single stable attractor and the baseline is not an artefact of the initial
condition.

## Check 3: harmaline pharmacokinetics against Fig. 7

Fig. 7A and 7B plot the model-predicted serum harmaline profiles the authors
themselves simulated. Peak values read off those (log-scale) panels are the
reference below; because they are digitised from a figure rather than printed,
they are compared at a deliberately loose tolerance.

```{r check-pk-harmaline}
pk_arms <- tidyr::crossing(dose = c(2, 5, 15), tg = c(0, 1))
pk_sims <- bind_rows(Map(
  function(dose, tg) {
    simulate_arm(paste0("HAR ", dose, " mg/kg"), dose_harmaline = dose,
                 tg = tg, tmax = 300, dt = 0.25)
  },
  pk_arms$dose, pk_arms$tg
))

cmax_tab <- pk_sims |>
  group_by(genotype, dose_harmaline) |>
  summarise(cmax_model = max(Cc_harmaline), .groups = "drop") |>
  mutate(cmax_figure = c(0.55, 2.5, 8.5, 0.55, 3.0, 11.0)) |>
  mutate(pct_diff = 100 * (cmax_model - cmax_figure) / cmax_figure)

stopifnot(max(abs(cmax_tab$pct_diff)) < 25)

cmax_tab |>
  rename(
    "Genotype" = genotype, "Harmaline dose (mg/kg)" = dose_harmaline,
    "Model Cmax (umol/L)" = cmax_model, "Fig. 7 Cmax (umol/L)" = cmax_figure,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = 2)
```

This is the check that settles the `L/min` versus `L/min/kg` ambiguity noted
under Units. Reading the two harmaline clearances as absolute L/min would put
these peaks roughly 30-fold lower than the panels the authors published, so the
per-kg reading is the only one consistent with the paper's own figure.

```{r fig-pk, echo = FALSE, fig.cap = "Replicates the shape of Figure 7A and 7B of Jiang 2016: simulated serum harmaline after 2, 5 and 15 mg/kg i.p."}
ggplot(pk_sims, aes(time, Cc_harmaline, colour = factor(dose_harmaline))) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~genotype) +
  scale_y_log10(limits = c(1e-3, 100)) +
  labs(x = "Time (min)", y = "Harmaline (umol/L)",
       colour = "Dose (mg/kg)") +
  theme_bw()
```

## Check 4: dose-proportionality and mass balance via PKNCA

Two quantitative claims in the Discussion (p. 501) are printed in the text, so
they are non-circular gates rather than figure readings:

* harmaline exposure rises "almost 20-fold" going from 2 to 15 mg/kg;
* early-time (0-60 min) 5-MeO-DMT exposure at 10 mg/kg rises "only by 20%" when
  the harmaline dose rises from 2 to 15 mg/kg.

The first is a direct test of the dose-dependent bioavailability: the dose
itself only rises 7.5-fold, so the rest must come from FH rising from 34.6% to
90.3%.

```{r check-nca-harmaline}
conc_har <- pk_sims |>
  filter(!is.na(Cc_harmaline)) |>
  transmute(id = paste(genotype, dose_harmaline), time, conc = Cc_harmaline,
            genotype, dose_harmaline)

dose_har <- conc_har |>
  group_by(id, genotype, dose_harmaline) |>
  summarise(time = 0, .groups = "drop")

o_conc <- PKNCA::PKNCAconc(conc_har, conc ~ time | id)
o_dose <- PKNCA::PKNCAdose(dose_har, ~ time | id)
res_har <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(start = 0, end = 300, auclast = TRUE, cmax = TRUE)
))

auc_har <- as.data.frame(res_har) |>
  filter(PPTESTCD == "auclast") |>
  left_join(distinct(conc_har, id, genotype, dose_harmaline), by = "id")

fold_wt <- with(
  auc_har[auc_har$genotype == "Wild-type", ],
  PPORRES[dose_harmaline == 15] / PPORRES[dose_harmaline == 2]
)

# Dose ratio 7.5 x bioavailability ratio (0.903 / 0.346 = 2.61) = 19.6.
stopifnot(fold_wt > 15, fold_wt < 24)
c(harmaline_auc_fold_2_to_15 = round(fold_wt, 1))
```

The model gives a `r round(fold_wt, 1)`-fold rise, matching the paper's "almost
20-fold". A flat-bioavailability encoding would give only 7.5-fold, so this gate
fails loudly if the FH lookup is wrong.

Mass balance closes the same loop from the other side: for a linear disposition,
`CL x AUC(0-inf)` must return the absorbed dose.

```{r check-mass-balance}
mb <- pk_sims |>
  filter(genotype == "Wild-type") |>
  group_by(dose_harmaline) |>
  summarise(
    # Trapezoid over the simulated grid plus the analytic terminal tail.
    auc_last = sum(diff(time) * (head(Cc_harmaline, -1) + tail(Cc_harmaline, -1)) / 2),
    c_last = tail(Cc_harmaline, 1),
    .groups = "drop"
  ) |>
  mutate(
    lambda_z = 0.0962 / (2.43 + 2.86), # terminal slope of the 2-cmt system
    auc_inf = auc_last + c_last / lambda_z,
    absorbed = dose_harmaline / MW_HARMALINE * 1000 *
      c(0.346, 0.742, 0.903),
    recovered = 0.0962 * auc_inf,
    pct_err = 100 * (recovered - absorbed) / absorbed
  )

stopifnot(max(abs(mb$pct_err)) < 5)
mb |>
  select(dose_harmaline, absorbed, recovered, pct_err) |>
  rename("Dose (mg/kg)" = dose_harmaline, "F x Dose (umol/kg)" = absorbed,
         "CL x AUCinf (umol/kg)" = recovered, "Error (%)" = pct_err) |>
  knitr::kable(digits = 2)
```

Now the second printed claim, which exercises the competitive-inhibition terms.

```{r check-ddi-exposure}
ddi_sims <- bind_rows(lapply(c(2, 5, 15), function(dh) {
  simulate_arm(paste0("HAR ", dh, " + MEO 10"), dose_harmaline = dh,
               dose_meodmt = 10, tg = 0, tmax = 300, dt = 0.25)
}))

auc_meo_60 <- ddi_sims |>
  filter(time >= 15, time <= 60) |>
  group_by(dose_harmaline) |>
  summarise(
    auc = sum(diff(time) * (head(Cc_meodmt, -1) + tail(Cc_meodmt, -1)) / 2),
    .groups = "drop"
  )

meo_fold <- with(auc_meo_60, auc[dose_harmaline == 15] / auc[dose_harmaline == 2])

# Paper: "only increased by 20%"; anything near 1 rather than a large factor.
stopifnot(meo_fold > 1.0, meo_fold < 1.5)
c(meodmt_early_auc_fold_2_to_15 = round(meo_fold, 2))
```

A `r round(meo_fold, 2)`-fold rise, against the paper's ~1.2. The point the
authors draw from this is that the *late* hyperpyrexia at toxic dose
combinations cannot be attributed to 5-MeO-DMT exposure, because 5-MeO-DMT
exposure barely moves while harmaline exposure moves 20-fold.

## Check 5: replicating the thermoregulatory figures

### Figure 2 -- saline only

```{r fig2}
fig2 <- bind_rows(
  bind_rows(lapply(c(0, 1), function(tg) {
    simulate_arm("Baseline", tg = tg, tmax = 300)
  })),
  bind_rows(lapply(c(0, 1), function(tg) {
    simulate_arm("Saline", tg = tg, tmax = 300, inj_extra = TRUE,
                 t_second = 0)
  })),
  bind_rows(lapply(c(0, 1), function(tg) {
    # Two handling events: one at 0 and one at 15 min.
    ev <- rxode2::et(seq(0, 300, by = 0.5), cmt = "temp")
    ev <- rxode2::et(ev, amt = 1, cmt = "stress", time = 0)
    ev <- rxode2::et(ev, amt = 1, cmt = "stress", time = 15)
    d <- as.data.frame(ev)
    d$CYP2D6_TG <- tg
    d$INJ_REPEAT <- 1L
    d$CONMED_HARMALINE <- 0L
    d$DOSE_HARMALINE_MGKG <- 0
    out <- rxode2::rxSolve(mod, d, returnType = "data.frame")
    out$arm <- "Saline + saline"
    out$genotype <- ifelse(tg == 1, "Tg-CYP2D6", "Wild-type")
    out
  }))
)

fig2_peaks <- fig2 |>
  group_by(arm) |>
  summarise(peak = max(temp), t_peak = time[which.max(temp)], .groups = "drop")

stopifnot(
  # Single saline peaks near 37.1 degC around 15 min (Fig. 2, dashed line).
  abs(fig2_peaks$peak[fig2_peaks$arm == "Saline"] - 37.1) < 0.4,
  # A second injection must raise the peak above a single one.
  fig2_peaks$peak[fig2_peaks$arm == "Saline + saline"] >
    fig2_peaks$peak[fig2_peaks$arm == "Saline"],
  # The undosed baseline never shows a transient.
  fig2_peaks$peak[fig2_peaks$arm == "Baseline"] < 36.4
)
knitr::kable(fig2_peaks, digits = 2)
```

```{r fig2-plot, echo = FALSE, fig.cap = "Replicates Figure 2 of Jiang 2016: CBT at baseline and after single or double saline injection."}
ggplot(fig2, aes(time, temp, colour = arm)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~genotype) +
  coord_cartesian(ylim = c(34, 40)) +
  labs(x = "Time (min)", y = "CBT (degC)", colour = NULL) +
  theme_bw()
```

### Figure 3 -- each drug alone

```{r fig3}
har_arms <- tidyr::crossing(dose = c(2, 5, 15), tg = c(0, 1))
fig3_har <- bind_rows(Map(
  function(dose, tg) {
    simulate_arm(paste0("HAR ", dose), dose_harmaline = dose, tg = tg,
                 tmax = 180)
  },
  har_arms$dose, har_arms$tg
))
meo_arms <- tidyr::crossing(dose = c(2, 10, 20), tg = c(0, 1))
fig3_meo <- bind_rows(Map(
  function(dose, tg) {
    simulate_arm(paste0("MEO ", dose), dose_meodmt = dose, tg = tg,
                 t_second = 0, tmax = 240)
  },
  meo_arms$dose, meo_arms$tg
))

meo_peaks <- fig3_meo |>
  group_by(genotype, dose_meodmt) |>
  summarise(peak = max(temp), .groups = "drop") |>
  mutate(fig3_peak = c(37.2, 37.55, 37.8, 37.2, 37.55, 37.8))

har_nadirs <- fig3_har |>
  group_by(genotype, dose_harmaline) |>
  summarise(nadir = min(temp), .groups = "drop")

stopifnot(
  # 5-MeO-DMT hyperthermia: digitised from Fig. 3C/D, so a loose bound.
  max(abs(meo_peaks$peak - meo_peaks$fig3_peak)) < 0.5,
  # Dose-ordered in both directions.
  all(diff(meo_peaks$peak[meo_peaks$genotype == "Wild-type"]) > 0),
  all(diff(har_nadirs$nadir[har_nadirs$genotype == "Wild-type"]) < 0),
  # The paper's central genotype finding: harmaline hypothermia is deeper in
  # wild-type mice, while 5-MeO-DMT hyperthermia is genotype-independent.
  har_nadirs$nadir[har_nadirs$genotype == "Wild-type" &
    har_nadirs$dose_harmaline == 15] <
    har_nadirs$nadir[har_nadirs$genotype == "Tg-CYP2D6" &
      har_nadirs$dose_harmaline == 15],
  max(abs(diff(meo_peaks$peak[meo_peaks$dose_meodmt == 20]))) < 0.1
)
knitr::kable(meo_peaks, digits = 2)
knitr::kable(har_nadirs, digits = 2)
```

```{r fig3-plot, echo = FALSE, fig.cap = "Replicates Figure 3 of Jiang 2016: harmaline-induced hypothermia (top) and 5-MeO-DMT-induced hyperthermia (bottom)."}
bind_rows(
  mutate(fig3_har, panel = "Harmaline alone", dose = dose_harmaline),
  mutate(fig3_meo, panel = "5-MeO-DMT alone", dose = dose_meodmt)
) |>
  ggplot(aes(time, temp, colour = factor(dose))) +
  geom_line(linewidth = 0.7) +
  facet_grid(panel ~ genotype) +
  coord_cartesian(ylim = c(32, 39)) +
  labs(x = "Time (min)", y = "CBT (degC)", colour = "Dose (mg/kg)") +
  theme_bw()
```

The wild-type nadir after 15 mg/kg harmaline is
`r round(har_nadirs$nadir[har_nadirs$genotype == "Wild-type" & har_nadirs$dose_harmaline == 15], 2)`
degC against roughly 33.3 degC for the fitted line in Fig. 3A -- see Errata.

### Figure 4 -- the biphasic combination response

Co-administering harmaline with 2 mg/kg 5-MeO-DMT is the paper's signature
result: hypothermia early (0-45 min), hyperthermia late (45-120 min).

```{r fig4}
fig4_arms <- tidyr::crossing(dose = c(2, 5, 15), tg = c(0, 1))
fig4 <- bind_rows(Map(
  function(dose, tg) {
    simulate_arm(paste0("HAR ", dose, " + MEO 2"), dose_harmaline = dose,
                 dose_meodmt = 2, tg = tg, tmax = 240)
  },
  fig4_arms$dose, fig4_arms$tg
))

biphasic <- fig4 |>
  group_by(genotype, dose_harmaline) |>
  summarise(
    early_peak = max(temp[time <= 20]),
    early_min = min(temp[time <= 45]),
    late_max = max(temp[time > 45 & time <= 120]),
    .groups = "drop"
  ) |>
  mutate(swing = late_max - early_min)

stopifnot(
  # The shape is genuinely biphasic: an early handling-driven rise, a fall, and
  # then a second, 5-MeO-DMT-driven peak.
  all(biphasic$early_peak > biphasic$early_min),
  all(biphasic$swing > 0.5),
  # More harmaline means a lower late peak: harmaline's hypothermic action
  # progressively offsets the 5-MeO-DMT hyperthermia. Strictly ordered in both
  # genotypes (WT 37.71 / 37.37 / 36.92; Tg 37.72 / 37.51 / 37.23).
  all(diff(biphasic$late_max[biphasic$genotype == "Wild-type"]) < 0),
  all(diff(biphasic$late_max[biphasic$genotype == "Tg-CYP2D6"]) < 0),
  # The same offset suppresses the early handling-driven peak
  # (WT 37.75 / 37.40 / 36.32; Tg 37.75 / 37.48 / 36.65).
  all(diff(biphasic$early_peak[biphasic$genotype == "Wild-type"]) < 0),
  all(diff(biphasic$early_peak[biphasic$genotype == "Tg-CYP2D6"]) < 0),
  # Genotype effect survives into the combination: at 15 mg/kg the wild-type
  # trough is the deeper one, because harmaline is cleared more slowly.
  biphasic$early_min[biphasic$genotype == "Wild-type" &
    biphasic$dose_harmaline == 15] <
    biphasic$early_min[biphasic$genotype == "Tg-CYP2D6" &
      biphasic$dose_harmaline == 15]
)
knitr::kable(biphasic, digits = 2)
```

Two features are worth reading off that table. First, the early-phase
hypothermia is shallow here, and that is the model behaving as the paper
describes rather than a defect: the combination arms use the repeat-handling
regimen, which more than halves `kS-H`, and they carry two handling events whose
stress signal pushes temperature up over the same window. Only the wild-type
15 mg/kg arm dips below its starting temperature. Second, at low harmaline doses
the *early*, handling-driven peak is the taller of the two; what rising
harmaline does is push both peaks down monotonically, which is the hypothermic
arm progressively offsetting both the stress and the 5-MeO-DMT hyperthermia.

```{r fig4-plot, echo = FALSE, fig.cap = "Replicates Figure 4E and 4F of Jiang 2016: harmaline plus 2 mg/kg 5-MeO-DMT."}
ggplot(fig4, aes(time, temp, colour = factor(dose_harmaline))) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~genotype) +
  coord_cartesian(ylim = c(33, 39)) +
  labs(x = "Time (min)", y = "CBT (degC)", colour = "Harmaline (mg/kg)") +
  theme_bw()
```

## Check 6: the 4-fold potency shift is wired to the right arm

`SC50-M` must take the DDI value only when harmaline is actually present.

```{r check-sc50}
ec50_nonddi <- exp(mod$theta[["lec50_meodmt_nonddi"]])
ec50_ddi <- exp(mod$theta[["lec50_meodmt_ddi"]])

# Same 5-MeO-DMT dose, with and without harmaline pretreatment.
with_har <- simulate_arm("with", dose_harmaline = 5, dose_meodmt = 2, tmax = 240)
no_har <- simulate_arm("without", dose_meodmt = 2, t_second = 15, tmax = 240)

stopifnot(
  abs(ec50_nonddi - 1.88) < 1e-6,
  abs(ec50_ddi - 0.496) < 1e-6,
  abs(ec50_nonddi / ec50_ddi - 3.79) < 0.02,
  # The transduction signal driving thermogenesis must be larger under DDI.
  max(with_har$transit3) > max(no_har$transit3)
)
c(
  SC50_nonDDI = ec50_nonddi, SC50_DDI = ec50_ddi,
  fold = round(ec50_nonddi / ec50_ddi, 2)
)
```

## Assumptions, deviations and errata

* **Clearance units.** `CLother-H` (0.0962) and `CLCYP2D6-H` (0.0608) are
  printed as "L/min" in Methods 2.4 while every neighbouring volume, clearance
  and Vmax carries "/kg". They are encoded here as **L/min/kg**. Check 3 is the
  evidence: the per-kg reading reproduces the peak concentrations in the
  authors' own Fig. 7A/B to within 5%, whereas an absolute-L/min reading would
  be roughly 30-fold low and could not produce the observed hypothermia at all.
  This is treated as a typographical omission in the source.

* **Form of the two inhibition terms.** The paper names `Ki(M)-H` and
  `Ki(D)-H` but never writes the inhibited rate equations. Fig. 1 draws both
  inhibition arrows terminating in a bar on the **Km** label of the respective
  pathway, which is competitive inhibition, so both are encoded as
  `Km_app = Km * (1 + Cc_harmaline / Ki)`. The "other murine" route carries no
  inhibition arrow and is left uninhibited.

* **Form of the CYP2D6 O-demethylation term.** Likewise unwritten in the text.
  The Fig. 1 schematic prints it explicitly as
  `Vmax(D)-M x (1 + fmCYP2D6(D))`, which is what the model implements.

* **Residual error.** Eq. (10) defines the variance model as
  `VAR = (sigma1 + sigma2 * Y)^2`, a combined additive-plus-proportional SD, but
  no numeric sigma is printed in Table 1 or anywhere else in the paper. The
  structure is preserved and both SDs are held at `fixed(0)`; a user re-fitting
  this model must supply their own starting values. (The printed Eq. (10) reads
  `(sigma1 + sigma1 * Y)` while the surrounding text defines both `sigma1` and
  `sigma2`, so the second subscript is itself a typo in the source.)

* **No inter-individual variability.** The fit was naive-pooled, so the paper
  reports no IIV and none is invented. Every simulation here is deterministic.

* **Bioavailability away from the published doses.** `FH` is published only at
  2, 5 and 15 mg/kg. The model reproduces those six values exactly and
  interpolates linearly between them, holding the nearest value outside the
  range. This is an encoding choice: the paper's own external-validation arm
  (Fig. 5A/B) uses 10 mg/kg harmaline, a dose for which it reports no `FH`, and
  does not say what it used.

* **Molecular weights.** 214.26 g/mol (harmaline) and 218.30 g/mol (5-MeO-DMT)
  are chemical constants used only to convert the published mg/kg doses into the
  model's umol/kg; they are not from the paper.

* **Deviation -- depth of harmaline hypothermia at 15 mg/kg.** The model reaches
  a wild-type nadir near 32.6 degC where the fitted line in Fig. 3A bottoms out
  around 33.3 degC (Tg-*CYP2D6*: 33.7 versus roughly 34.6 degC). Direction,
  timing and the genotype ordering are all reproduced, and the lower doses agree
  closely; the discrepancy is confined to the highest harmaline dose. Since the
  PK at that dose matches Fig. 7A to within 5%, the residual difference sits in
  the PD arm. It is recorded here rather than tuned away, and is excluded from
  the assertions above.

* **Not reproduced -- the toxic dose combinations of Fig. 6.** The authors state
  plainly that their model fails there, underestimating the early rise and
  overestimating the late fall in CBT for 5 or 15 mg/kg harmaline plus 10 mg/kg
  5-MeO-DMT. That is a published limitation of the model, not of this encoding,
  and no attempt is made to match those profiles.

* **Upstream PK source.** All PK parameters are fixed values carried over from
  the companion PK-interaction study (Jiang 2013, *Drug Metab Dispos*
  41:975-86), and every one of them is reprinted in Methods 2.4 of this paper,
  which is where they were read from. The upstream paper itself is not open
  access; the structural forms it would have supplied are instead resolved from
  Fig. 1 as described above.

* **Known convention warning -- the `temp` observation variable.**
  `checkModelConventions()` emits one warning for this model: the single-output
  observation variable `temp` is not a registered canonical. That is deliberate.
  The standing convention is that a *compartment* canonical is promoted only
  once a second, independent paper uses the same state, so core body temperature
  is carried here in `paper_specific_compartments` rather than minted as a
  library-wide canonical on the strength of one paper -- particularly since
  `temp` is a collision-prone token (a future model may want it for something
  else, and the paper's own abbreviation `CBT` is an equally good candidate).
  The warning is the correct signal for a first sighting; the trigger to revisit
  is a second thermoregulation PK/PD model arriving in the library, at which
  point the state should be registered in
  `inst/references/compartment-names.md` and this note removed. All other
  convention checks pass, and the four covariate columns this model introduces
  (`CYP2D6_TG`, `INJ_REPEAT`, `CONMED_HARMALINE`, `DOSE_HARMALINE_MGKG`) are
  registered in `inst/references/covariate-columns.md`.
