---
title: "Sirolimus (Jiao 2009)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sirolimus (Jiao 2009)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Jiao_2009_sirolimus"))$reference`
* Description: `r rxode2::rxode(readModelDb("Jiao_2009_sirolimus"))$description`
* Article: <https://doi.org/10.1111/j.1365-2125.2009.03392.x>

Jiao, Shi, Li, and Zhong (Huashan Hospital, Fudan University) developed a
one-compartment population PK model for oral sirolimus from 804 trough
concentrations collected in 112 Chinese adult de novo renal transplant
recipients enrolled at five clinical centres (Methods + Results). All
patients received induction therapy with ciclosporin (CsA), sirolimus, and
corticosteroids during the first three months, followed by either CsA dose
reduction (Phase I) or CsA discontinuation (Phase II). Because the sampling
design was trough-only, the absorption rate ka was held at the literature
value 0.752 1/h (Zahir et al. 2006, paper ref [32]) and apparent clearance
(CL/F) and apparent volume of distribution (V/F) were the estimated
parameters. After forward stepwise inclusion and backward elimination, five
covariates were retained on CL/F (total cholesterol, CsA trough
concentration, sirolimus daily dose, silymarin co-therapy, glycyrrhizin
co-therapy) and one on V/F (CsA trough concentration). The model is the
first population PK characterisation of sirolimus in Chinese renal
transplant patients and is notable for identifying silymarin (milk thistle)
and glycyrrhizin (licorice) co-therapy as novel covariates on sirolimus
CL/F in hepatically impaired patients.

# Population

The model-development cohort comprised 112 Chinese adult de novo primary
renal allograft recipients (78 male / 34 female; 30.4 percent female)
enrolled across five clinical centres in Guangzhou, Shanghai, and Beijing
between June 2004 and August 2006 (Methods, Study design; Table 1). Mean
age was 42 years (range 19-68), mean body weight 60.4 kg (range 44.5-91.0),
mean body surface area 1.70 m^2 (range 1.33-2.18). The cohort split between
Phase I (n = 71, CsA dose reduction) and Phase II (n = 41, CsA
discontinuation), with most allografts sourced from cadaveric donors (104
of 112, 93 percent). Sirolimus was initiated at a loading dose of 6 mg PO
within 24 h of transplant surgery, followed by a maintenance dose of 0.5-6
mg/day titrated to whole-blood trough targets (6-12 ng/mL months 1-3,
12-20 ng/mL months 4-6 of Phase II, 10-20 ng/mL months 7-12 of Phase II).
The trial is registered as NCT00257387 and was sponsored by Wyeth
Pharmaceutical Co., Ltd., China.

Bioanalysis used three external laboratories with two HPLC-UV methods and
one LC-MS/MS method, with all three validated through the International
Sirolimus Proficiency Testing Scheme. CsA whole-blood concentrations were
measured by automated fluorescence polarisation with monoclonal antibodies
(AxSYM at Shanghai sites, TDx at the other three sites); no assay
cross-normalisation was applied to the original C0 values in the final
model (sensitivity analysis showed assay-rescaling shifted CsA-related
parameters by < 20 percent with < 2 OFV units alternation).

The same information is available programmatically via the model's
`population` metadata.

```{r population-meta}
mod <- readModelDb("Jiao_2009_sirolimus")
str(rxode2::rxode(mod)$population)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Jiao_2009_sirolimus.R`. The
table below collects the same information in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (typical CL/F at cohort-median covariates, L/h) | 10.1 | Table 2 theta1 = 10.1 L/h (3.0% RSE) |
| `lvc` (typical V/F at CsA-trough reference, L) | 3670 | Table 2 theta7 = 3670 L (9.3% RSE) |
| `lka` (ka, 1/h; literature-fixed) | 0.752 | Methods + paper ref [32] (Zahir 2006); ka fixed, IIV not estimated |
| `e_tchol_cl` (linear-deviation effect of TCHOL on CL/F, L/h per mmol/L) | -0.662 | Table 2 theta2 = 0.662 (25.1% RSE); minus sign per Eq. 9 bracket |
| `e_cp_csa_ngml_cl` (linear-deviation effect of CsA C0 on CL/F, L/h per ng/mL) | -0.00417 | Table 2 theta3 = 4.17e-3 (44.4% RSE); minus sign per Eq. 9 |
| `e_conmed_silymarin_cl` (log effect of silymarin on CL/F) | log(0.650) | Table 2 theta4 = 0.650 (16.3% RSE); 35% reduction in CL/F |
| `e_conmed_glycyrrhizin_cl` (log effect of glycyrrhizin on CL/F) | log(0.661) | Table 2 theta5 = 0.661 (25.4% RSE); 34% reduction in CL/F |
| `e_dose_cl` (power exponent of (DOSE/2) on CL/F) | 0.479 | Table 2 theta6 = 0.479 (16.6% RSE) |
| `e_cp_csa_ngml_vc` (linear-deviation effect of CsA C0 on V/F, L per ng/mL) | -7.27 | Table 2 theta8 = 7.27 (14.9% RSE); minus sign per Eq. 10 |
| `etalcl` (omega^2 on log CL/F) | 0.0551 | Table 2 omega_CL/F = 23.8% (26.7% RSE); omega^2 = log(1 + 0.238^2) |
| `etalvc` (omega^2 on log V/F) | 0.2789 | Table 2 omega_V/F = 56.7% (24.8% RSE); omega^2 = log(1 + 0.567^2) |
| `expSd` (log-normal residual SD) | 0.299 | Table 2 sigma = 29.9% (8.0% RSE); exponential error model per Methods Eq. 3 |
| Final equation: `CL/F = [theta1 - theta2*(TC - 5.66) - theta3*(C0 - 104)] * theta4^SLM * theta5^GLZ * (DDS/2)^theta6` | n/a | Eq. 9 (page 52); also reproduced in Table 2 footer |
| Final equation: `V/F = theta7 - theta8*(C0 - 104)` | n/a | Eq. 10 (page 52); also reproduced in Table 2 footer |
| Cohort medians used as covariate references (TC = 5.66, C0 = 104, DDS = 2) | n/a | Table 1: TC median 5.66 mmol/L, CsA C0 median 104 ng/mL, sirolimus DDS median 2 mg/day |
| Structural model: 1-compartment with first-order absorption (NONMEM ADVAN2/TRANS2) | n/a | Methods, Population pharmacokinetic analysis |
| IIV form: log-normal exponential (Pi = P_typ * exp(eta_i)) | n/a | Methods Eq. 1 |
| Residual model: exponential (Y = IPRED * exp(eps); chosen over additive and combined per OFV / GOF) | n/a | Methods Eq. 3 + Results, Model building and evaluations |

# Virtual cohort

Original observed concentration data are not publicly available. To verify
the packaged model reproduces the paper's key quantitative findings, the
simulation below builds four illustrative cohorts that hold the
non-targeted covariates at cohort medians while varying the covariate of
interest:

* **Reference**: all covariates at cohort median (TCHOL = 5.66 mmol/L,
  CsA C0 = 104 ng/mL, daily dose = 2 mg, no silymarin / glycyrrhizin
  co-therapy). Typical CL/F = 10.1 L/h by construction (Eq. 9).
* **Silymarin co-therapy**: as Reference plus `CONMED_SILYMARIN = 1`,
  expected to reduce typical CL/F by 35 percent to 6.6 L/h.
* **Glycyrrhizin co-therapy**: as Reference plus `CONMED_GLYCYRRHIZIN = 1`,
  expected to reduce typical CL/F by 34 percent to 6.7 L/h.
* **High daily dose**: as Reference but daily dose = 4 mg/day, expected to
  increase typical CL/F by a factor of (4/2)^0.479 = 1.39 to 14.1 L/h.

```{r cohort}
set.seed(20090401L)

n_per_group <- 60L

# Per-subject IIV is generated by rxSolve; the covariates below are fixed
# per subject so the cohort-level effect on typical CL/F is interpretable.
make_cohort <- function(label, n, dose_mg = 2, csa = 104, slm = 0L, glz = 0L,
                        tchol = 5.66, id_offset = 0L) {
  ids <- id_offset + seq_len(n)
  # Sirolimus has a long apparent half-life (V/F approximately 3670 L,
  # CL/F approximately 10 L/h -> t1/2 approximately 250 h or 10 d), so a
  # 60-day daily-dose regimen is needed to approach steady state. We
  # observe sparse troughs at day 30 and 60, then dense observations
  # spanning the final dosing interval (days 59-60) for NCA.
  dose_times <- seq(0, by = 24, length.out = 60)  # 60 daily doses
  obs_times  <- sort(unique(c(
    30 * 24,                                       # day-30 trough
    59 * 24 + c(0, 1, 2, 4, 6, 8, 12, 18, 24)      # final-day dense
  )))
  dose_rows <- tidyr::expand_grid(id = ids, time = dose_times) |>
    dplyr::mutate(evid = 1L, amt = dose_mg, cmt = "depot")
  obs_rows  <- tidyr::expand_grid(id = ids, time = obs_times) |>
    dplyr::mutate(evid = 0L, amt = 0, cmt = "central")
  dplyr::bind_rows(dose_rows, obs_rows) |>
    dplyr::mutate(
      TCHOL                = tchol,
      CP_CSA_NGML          = csa,
      DOSE                 = dose_mg,
      CONMED_SILYMARIN     = slm,
      CONMED_GLYCYRRHIZIN  = glz,
      cohort               = label
    ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  make_cohort("Reference (DDS 2 mg/day, no herbs)",
              n_per_group, id_offset = 0L),
  make_cohort("Silymarin co-therapy",
              n_per_group, slm = 1L, id_offset = 1L * n_per_group),
  make_cohort("Glycyrrhizin co-therapy",
              n_per_group, glz = 1L, id_offset = 2L * n_per_group),
  make_cohort("High dose (DDS 4 mg/day)",
              n_per_group, dose_mg = 4, id_offset = 3L * n_per_group)
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))

events$cohort <- factor(
  events$cohort,
  levels = c(
    "Reference (DDS 2 mg/day, no herbs)",
    "Silymarin co-therapy",
    "Glycyrrhizin co-therapy",
    "High dose (DDS 4 mg/day)"
  )
)
```

# Simulation

The full simulation carries between-subject variability (the IIV reported
in Table 2). The deterministic typical-value check that follows uses
`rxode2::zeroRe()` to confirm the encoded equation matches Eq. 9 exactly
at the worked-out reference values.

```{r simulate}
sim <- rxode2::rxSolve(
  mod,
  events = as.data.frame(events),
  keep   = c("cohort", "TCHOL", "CP_CSA_NGML", "DOSE",
             "CONMED_SILYMARIN", "CONMED_GLYCYRRHIZIN")
) |>
  as.data.frame()
```

# Replicate published equations

## Typical CL/F at the worked-equation references (Eq. 9, Table 2)

The most direct quantitative check of the encoding is the typical CL/F at
the cohort-median covariates and at each single-covariate perturbation.
Because the bracketed term in Eq. 9 is centred on the cohort medians and
the herb / dose multipliers default to 1 at the reference, the typical
CL/F at all-median covariates must equal theta1 = 10.1 L/h.

```{r typical-cl}
mod_typical <- rxode2::zeroRe(mod)

scenarios <- tibble::tribble(
  ~scenario,                                ~TCHOL, ~CP_CSA_NGML, ~DOSE, ~CONMED_SILYMARIN, ~CONMED_GLYCYRRHIZIN, ~paper_cl,
  "Cohort medians (Table 2 theta1)",          5.66,         104,     2,                0L,                   0L,      10.10,
  "Silymarin co-therapy",                     5.66,         104,     2,                1L,                   0L,      10.10 * 0.650,
  "Glycyrrhizin co-therapy",                  5.66,         104,     2,                0L,                   1L,      10.10 * 0.661,
  "DDS 4 mg/day",                             5.66,         104,     4,                0L,                   0L,      10.10 * (4 / 2)^0.479,
  "TCHOL 8 mmol/L (high)",                    8.00,         104,     2,                0L,                   0L,      10.10 + (-0.662) * (8 - 5.66),
  "CsA C0 200 ng/mL (high)",                  5.66,         200,     2,                0L,                   0L,      10.10 + (-0.00417) * (200 - 104)
)

# Build a one-record-per-scenario event table with a single dose and a single
# observation immediately after, just to force rxode2 to evaluate the typical
# CL/F via the algebraic intermediates the model reports. rxode2 requires the
# standard event columns (id, time, evid, amt, cmt) to precede covariate
# columns in the input data frame, so we build with that column order
# explicitly.
scenario_events <- scenarios |>
  dplyr::mutate(id = dplyr::row_number()) |>
  dplyr::select(id, scenario, TCHOL, CP_CSA_NGML, DOSE,
                CONMED_SILYMARIN, CONMED_GLYCYRRHIZIN)

dose_rows <- scenario_events |>
  dplyr::mutate(time = 0, evid = 1L, amt = ifelse(DOSE > 0, DOSE, 2),
                cmt = "depot")
obs_rows  <- scenario_events |>
  dplyr::mutate(time = 0.01, evid = 0L, amt = 0, cmt = "central")

scen_events <- dplyr::bind_rows(dose_rows, obs_rows) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  dplyr::select(id, time, evid, amt, cmt,
                scenario, TCHOL, CP_CSA_NGML, DOSE,
                CONMED_SILYMARIN, CONMED_GLYCYRRHIZIN)

scen_sim <- rxode2::rxSolve(
  mod_typical,
  events = as.data.frame(scen_events),
  keep   = c("scenario", "TCHOL", "CP_CSA_NGML", "DOSE",
             "CONMED_SILYMARIN", "CONMED_GLYCYRRHIZIN")
) |>
  as.data.frame()

scen_cl <- scen_sim |>
  dplyr::filter(time > 0) |>
  dplyr::select(scenario, cl_model = cl) |>
  dplyr::left_join(
    scenarios |> dplyr::select(scenario, cl_paper = paper_cl),
    by = "scenario"
  ) |>
  dplyr::mutate(diff_pct = 100 * (cl_model - cl_paper) / cl_paper)

knitr::kable(
  scen_cl,
  digits  = c(0, 3, 3, 2),
  col.names = c("Scenario", "CL/F model (L/h)", "CL/F paper (L/h)", "Diff (%)"),
  caption = paste(
    "Typical CL/F predicted by the packaged model versus the value computed",
    "directly from Jiao 2009 Eq. 9 at each scenario. Differences should be at",
    "the numerical-noise level (< 0.01%)."
  )
)
```

A match within numerical noise across all six scenarios confirms the
covariate equation is encoded faithfully. The two herb scenarios reproduce
the 35 percent and 34 percent CL/F reductions reported in the Discussion;
the DDS = 4 mg/day case shows the (4 / 2)^0.479 = 1.39-fold increase
attributable to the nonlinear self-dose covariate; and the high-TCHOL /
high-CsA cases show the expected decreases.

## Distribution of trough concentrations at steady state

The paper's median observed sirolimus trough concentration is 8.25 ng/mL
(Table 1, range 1.50-30.6 ng/mL across 804 observations). At cohort
median covariates, the model predicts a steady-state Cavg over the 24 h
dosing interval of `dose / (CL/F * tau) * 1000 = 2 / (10.1 * 24) * 1000 =
8.25 ng/mL` exactly, so the cohort-median trough at steady state should
land in the same neighbourhood. Because the sirolimus apparent half-life
in this model is roughly 250 h (V/F = 3670 L, CL/F = 10.1 L/h), steady
state is approached over 5 half-lives = approximately 50 days; the day-59
trough below is therefore a reasonable steady-state surrogate.

```{r trough-summary}
trough <- sim |>
  dplyr::filter(!is.na(Cc), abs(time - 30 * 24) < 1e-6 |
                              abs(time - 59 * 24) < 1e-6) |>
  dplyr::mutate(day = round(time / 24)) |>
  dplyr::group_by(cohort, day) |>
  dplyr::summarise(
    median = median(Cc, na.rm = TRUE),
    q05    = quantile(Cc, 0.05, na.rm = TRUE),
    q95    = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(
  trough,
  digits  = 2,
  caption = paste(
    "Simulated trough Cc (ng/mL) at day 30 and day 59 across the four illustrative",
    "cohorts. Jiao 2009 Table 1 reports a median observed trough of 8.25 ng/mL",
    "(range 1.50-30.6) across the full TDM dataset; the reference-cohort day-59",
    "trough should land near 8 ng/mL with the herb cohorts shifted upward",
    "(lower CL/F) and the high-dose cohort shifted upward by accumulation."
  )
)
```

```{r trough-plot}
sim |>
  dplyr::filter(!is.na(Cc), abs(time - 59 * 24) < 1e-6) |>
  ggplot(aes(cohort, Cc)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  geom_hline(yintercept = c(6, 12), colour = "firebrick", linetype = "dashed") +
  geom_hline(yintercept = 8.25, colour = "darkgreen", linetype = "solid",
             linewidth = 0.4) +
  scale_y_log10() +
  labs(x = NULL, y = "Day-59 trough Cc (ng/mL)",
       title = "Steady-state trough by cohort at the cohort-median DDS = 2 mg/day",
       caption = paste(
         "Dashed red lines = months-1-3 / Phase-I TDM target band (6-12 ng/mL);",
         "solid green line = paper's median observed trough 8.25 ng/mL."
       )) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

The reference cohort centres on the paper's observed median trough; the
silymarin and glycyrrhizin co-therapy cohorts shift upward by roughly
1 / 0.65 to 1 / 0.66 (1.54-fold), consistent with the 35 / 34 percent
CL/F reductions reported in the Discussion; the high-dose cohort shifts
upward by the dose-accumulation ratio relative to the increase in CL/F
predicted by (DDS / 2)^0.479.

## Single-day concentration-time profile at steady state

```{r profile}
sim |>
  dplyr::filter(!is.na(Cc), time >= 59 * 24, time <= 60 * 24) |>
  dplyr::mutate(time_in_day = time - 59 * 24) |>
  dplyr::group_by(cohort, time_in_day) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time_in_day, Q50)) +
  geom_ribbon(aes(ymin = pmax(Q05, 1e-3), ymax = Q95), alpha = 0.25,
              fill = "steelblue") +
  geom_line(colour = "steelblue") +
  facet_wrap(~ cohort, ncol = 2) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 18, 24)) +
  labs(x = "Time within final dosing interval (h)", y = "Cc (ng/mL)",
       title = "Day-59 to day-60 concentration profile by cohort",
       caption = "Median (line) and 5-95 percent prediction interval (ribbon).")
```

# PKNCA validation

The paper does not report formal NCA on its trough-only TDM dataset, so
the validation target is the relationship `AUC0-tau,ss = dose / CL/F`
computed at each cohort's typical CL/F. PKNCA on the day-59 dosing
interval should reproduce these target values within numerical noise for
the typical-value (`zeroRe`) simulation.

```{r pknca}
ss_start <- 59 * 24
ss_end   <- 60 * 24

# Build the typical-value (no IIV) simulation explicitly so the
# comparison is anchored to the published equation rather than the
# IIV-sampled value.
sim_typ <- rxode2::rxSolve(
  mod_typical,
  events = as.data.frame(events),
  keep   = c("cohort", "TCHOL", "CP_CSA_NGML", "DOSE",
             "CONMED_SILYMARIN", "CONMED_GLYCYRRHIZIN")
) |>
  as.data.frame()

# Restrict to the final dosing interval and keep only one representative
# subject per cohort (typical-value -> all subjects in a cohort coincide).
conc_ss <- sim_typ |>
  dplyr::filter(!is.na(Cc), time >= ss_start, time <= ss_end) |>
  dplyr::group_by(cohort) |>
  dplyr::filter(id == min(id)) |>
  dplyr::ungroup() |>
  dplyr::select(id, time, Cc, cohort)

# Add a defensive time = ss_start row with Cc carried forward at the
# cohort's typical Cmin (PKNCA needs a row anchoring the interval start).
conc_ss_floor <- conc_ss |>
  dplyr::group_by(id, cohort) |>
  dplyr::summarise(time = ss_start, Cc = Cc[which.min(time)], .groups = "drop")
conc_ss <- dplyr::bind_rows(conc_ss_floor, conc_ss) |>
  dplyr::distinct(id, cohort, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

dose_ss <- events |>
  dplyr::filter(evid == 1L, time == ss_start) |>
  dplyr::group_by(cohort) |>
  dplyr::filter(id == min(id)) |>
  dplyr::ungroup() |>
  dplyr::select(id, time, amt, cohort)

conc_obj <- PKNCA::PKNCAconc(
  as.data.frame(conc_ss),
  Cc ~ time | cohort + id,
  concu = "ng/mL", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  as.data.frame(dose_ss),
  amt ~ time | cohort + id,
  doseu = "mg"
)

intervals <- data.frame(
  start    = ss_start,
  end      = ss_end,
  cmax     = TRUE,
  tmax     = TRUE,
  cmin     = TRUE,
  auclast  = TRUE,
  cav      = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- suppressWarnings(PKNCA::pk.nca(nca_data))

nca_tbl <- as.data.frame(nca_res$result) |>
  dplyr::select(cohort, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

# Compute the paper-equation expectations:
#   AUC0-tau,ss (ng*h/mL) = dose (mg) / CL/F (L/h) * 1000 (mg -> ng / L -> mL ratio)
#   Cavg,ss (ng/mL)       = AUC0-tau,ss / 24
target <- tibble::tribble(
  ~cohort,                                ~dose_mg, ~cl_paper,
  "Reference (DDS 2 mg/day, no herbs)",         2,      10.10,
  "Silymarin co-therapy",                       2,      10.10 * 0.650,
  "Glycyrrhizin co-therapy",                    2,      10.10 * 0.661,
  "High dose (DDS 4 mg/day)",                   4,      10.10 * (4 / 2)^0.479
) |>
  dplyr::mutate(
    auc_target  = dose_mg / cl_paper * 1000,
    cavg_target = auc_target / 24
  )

cmp <- nca_tbl |>
  dplyr::left_join(target, by = "cohort") |>
  dplyr::mutate(
    auc_diff_pct  = 100 * (auclast - auc_target) / auc_target,
    cavg_diff_pct = 100 * (cav - cavg_target) / cavg_target
  ) |>
  dplyr::select(cohort, cmax, cmin, tmax, auclast, auc_target, auc_diff_pct,
                cav, cavg_target, cavg_diff_pct)

knitr::kable(
  cmp,
  digits  = c(0, 2, 2, 1, 1, 1, 2, 2, 2, 2),
  col.names = c("Cohort", "Cmax,ss", "Cmin,ss", "Tmax",
                "AUC0-tau (sim)", "AUC0-tau (paper)", "AUC diff (%)",
                "Cavg (sim)", "Cavg (paper)", "Cavg diff (%)"),
  caption = paste(
    "Day-59 steady-state PKNCA on the typical-value (zeroRe) simulation.",
    "Cmax,ss / Cmin,ss / Tmax are absolute values from the simulation;",
    "AUC0-tau and Cavg are compared against the Eq. 9 prediction",
    "AUC0-tau,ss = dose * 1000 / CL/F. Differences should be at the",
    "numerical-integration tolerance (< 1%) once steady state is reached."
  )
)
```

Residual non-zero `AUC diff (%)` values, when present, reflect the small
gap between approximate steady state at day 59 and the analytical
steady-state assumption AUC0-tau,ss = dose / CL/F. Extending the dosing
horizon further closes the gap but is unnecessary for the validation goal.

# Assumptions and deviations

* **Equation 9 versus the paper's worked example.** Section "Model
  building and evaluations" reports an illustrative computation: "the CL/F
  of sirolimus can be estimated at 7.58 l h-1 for hypothetical patients
  with a C0 of 100 ng ml-1, TC of 5.0 mmol l-1, DDS of 3 mg and no
  co-administration with SLM or GLZ formulation, whereas CL/F is estimated
  at 4.93 l h-1 for the same patient with TC of 5.0 mmol l-1, C0 of
  100 ng ml-1, DDS of 3 mg, but with hepatic impairment and SLM
  co-therapy". The internal SLM ratio (4.93 / 7.58 = 0.650 = theta4) is
  consistent with Table 2, but a forward evaluation of Eq. 9 at the same
  covariate values gives a typical CL/F of 12.81 L/h (bracket = 10.554
  L/h * (3/2)^0.479 = 12.81 L/h), not 7.58 L/h. The discrepancy appears
  to be an arithmetic error in the paper's worked example rather than a
  problem with Eq. 9 itself: at cohort-median covariates the equation
  returns the reported typical CL/F of 10.1 L/h exactly, and the herb /
  CsA / TCHOL effect magnitudes match the Discussion narrative. The
  packaged model implements Eq. 9 verbatim. Downstream users who want to
  reproduce the published worked example will not obtain 7.58 L/h; they
  will obtain 12.81 L/h, which is what the equation evaluates to.
* **Sign of the bracketed deviation terms.** Eq. 9 writes the TCHOL and
  CsA C0 effects with explicit minus signs in the bracket (`10.1 -
  0.662 * (TC - 5.66) - 0.00417 * (C0 - 104)`). The model stores
  `e_tchol_cl = -0.662` and `e_cp_csa_ngml_cl = -0.00417` (signed
  coefficients) and writes the bracket with plus signs
  (`exp(lcl) + e_tchol_cl * (TCHOL - 5.66) + ...`); the arithmetic is
  identical to the paper.
* **Bracket positivity.** The linear-deviation bracket can in principle
  go negative at extreme covariate combinations, but evaluated across
  the cohort ranges (TCHOL 0.82 to 10.9, CsA C0 0 to 508) it stays
  positive: the most extreme realistic combination
  (`10.1 - 0.662 * (10.9 - 5.66) - 0.00417 * (508 - 104) = 4.95 L/h`)
  is still well above zero. Similarly, V/F = `3670 - 7.27 * (C0 - 104)`
  stays positive up to CsA C0 approximately 605 ng/mL, beyond the
  observed cohort maximum of 508. Users simulating extrapolated
  scenarios should check the typical-value prediction for sign-flip
  artefacts.
* **Apparent half-life is artificially long.** The paper's Discussion
  explicitly flags that V/F = 3670 L is "much higher than estimates of
  previous reports, which ranged from 117 to 1350 L" and attributes the
  inflated V/F to the trough-only sampling design ("trough concentrations
  alone were not adequate to estimate V/F"). The implied terminal-phase
  half-life t1/2 = ln(2) * V/F / CL/F = 0.693 * 3670 / 10.1 = 252 h is
  similarly inflated relative to the literature consensus near 60 h.
  The typical CL/F = 10.1 L/h, and hence the steady-state Cavg per dose
  level, is more reliable than the V/F-derived terminal slope; the
  steady-state Cavg = dose / (CL/F * tau) matches the cohort-median
  trough of 8.25 ng/mL exactly at the median 2 mg/day dose.
* **Ka fixed at 0.752 1/h (Zahir 2006).** Methods state ka was held at the
  literature value because no absorption-phase data were collected. The
  Results note that varying ka by 5- and 10-fold changed final estimates
  by no more than 10 and 4 percent respectively. The model file encodes
  ka with `fixed()` so downstream re-fits inherit the same fix; un-wrap
  if you want to re-estimate on a dataset with richer absorption
  sampling.
* **Bioavailability folded into apparent parameters.** All clearance and
  volume terms are "apparent" (X/F). The model uses `F = 1` as the
  structural anchor; the absolute bioavailability of oral sirolimus is
  approximately 17 percent per the Discussion but no absolute volume or
  clearance is identifiable from this dataset.
* **IIV correlation structure.** Methods Eq. 1 and the Results state the
  paper used a diagonal exponential IIV model with no off-block
  correlation between etas. The encoded model uses independent
  log-normal etas on `lcl` and `lvc`.
* **Residual error: exponential.** Methods Eq. 3 and the Results
  paragraph "Based on the minimum OFV and the distribution of residuals
  in the diagnostic plots of the base model, an exponential error model
  was selected" identify the residual structure. nlmixr2's
  `~ lnorm(expSd)` implements `Y ~ lnorm(IPRED, expSd)` on the
  log-concentration scale; the reported "29.9%" is the SD of the
  log-residual eps and maps directly to `expSd = 0.299`.
* **Herb / hepatic-impairment confounding.** Discussion notes "All
  patients with SLM or GLZ co-therapy had abnormal ALT values" and
  argues the cohort-level CL/F reduction reflects the combined effect
  of the herb and hepatic dysfunction. The packaged model encodes the
  herb indicator as the operative covariate (matching the paper's final
  equation), but downstream users should be aware that the effect size
  cannot be cleanly separated from the underlying liver dysfunction in
  this design.
* **Race / ethnicity.** All 112 subjects were Chinese adults. The paper
  notes typical CL/F = 10.1 L/h "lies within the range of previous
  studies from 8.9 to 14.1 L/h in White renal transplant patients", so
  there is no strong ethnicity-specific signal; the model can serve as
  a general adult-renal-transplant reference. `population$race_ethnicity`
  records the Chinese-only enrolment.
* **Trough-only sampling.** The 804 concentrations are end-of-dosing-
  interval troughs with a median 24 h (range 22.8-25 h) post-dose. The
  apparent-CL/F estimate is well-identified by trough density;
  absorption-phase and distribution-phase parameters are not.
* **Errata.** A search of the Wiley journal landing page for the article
  (Br J Clin Pharmacol 68:1, doi:10.1111/j.1365-2125.2009.03392.x) and
  a Google Scholar pass on the article title returned no errata or
  corrigenda at the time of extraction. If a future correction surfaces
  it should be encoded inline alongside Eq. 9 / Eq. 10 in the model
  file.
