---
title: "Benralizumab (Jin 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benralizumab (Jin 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r libraries, message = FALSE}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(ggplot2)
```

Jin 2025 updated the global benralizumab population pharmacokinetic (popPK) and
exposure-response (ER) models with data from three additional clinical studies,
two of them conducted in China or Asia, in order to characterise benralizumab
exposure and effectiveness in Chinese participants and to support extrapolation
to Chinese adolescents and children.

The paper contains **three separately fitted models**, and this package carries
one file for each:

| Model file | Endpoint | Structure |
|---|---|---|
| `Jin_2025_benralizumab` | Serum benralizumab | Two-compartment, first-order SC absorption |
| `Jin_2025_benralizumab_aaer` | Asthma exacerbation count | popPK layer + Poisson count with a concentration-driven Emax |
| `Jin_2025_benralizumab_fev1` | Pre-bronchodilator FEV1 | Baseline + exponential-onset placebo + **time**-driven Emax |

The FEV1 model carries no PK layer. Because the FEV1 exposure-response
relationship was flat across the studied dose range, the authors re-expressed
the treatment effect as a function of time rather than of concentration, so
that model is exposure-independent and valid only for the 30 mg subcutaneous
regimens studied.

```{r load-models}
mod_pk   <- rxode2::rxode(nlmixr2lib::readModelDb("Jin_2025_benralizumab"))
mod_aaer <- rxode2::rxode(nlmixr2lib::readModelDb("Jin_2025_benralizumab_aaer"))
mod_fev1 <- rxode2::rxode(nlmixr2lib::readModelDb("Jin_2025_benralizumab_fev1"))
```

# Population

```{r population}
pop <- mod_pk$population
str(pop, max.level = 1, give.attr = FALSE)
```

The popPK analysis pooled 17,465 serum or plasma benralizumab concentrations
from 2855 participants across 12 phase I-III studies: five phase I/II trials in
adults with asthma, four phase III trials in adults and adolescents with severe
asthma, and three additional studies added by this analysis (the phase I AMES
device-comparison trial in 180 healthy volunteers, a phase I trial in 36 healthy
Han Chinese volunteers, and the phase III MIRACLE trial in 695 Asian patients).

Jin 2025 Table 2 gives the baseline characteristics. The features that matter
most for the simulations below:

* Chinese participants were 281 of 2855 (10.0%) and Asian participants 590
  (20.7%). All Chinese participants are Asian by definition, so the two
  covariates are strongly correlated (Phi = 64.2%).
* Mean body weight was 14.5 kg lower in Chinese (64.1 kg) than in non-Chinese
  (78.6 kg) participants; Resource 6 gives 64.6 kg for Asian versus 80.4 kg for
  non-Asian participants.
* Only 58 participants were adolescents, of whom **2 were Asian and none were
  Chinese**, and **no participant was younger than 12 years**. Every Chinese
  adolescent and paediatric conclusion in the paper therefore rests on
  simulation rather than on observed data, which is the main reason the authors
  ran the external-validation step first.

# Source trace

Every value in the three model files comes from the electronic supplementary
material (ESM), whose sections the paper labels "Resource 1" to "Resource 19".

| Model quantity | Source location |
|---|---|
| popPK structure (2-compartment, first-order SC absorption) | Jin 2025 Methods 2.4; parameter names in Resource 10 |
| Covariate functional forms (power for continuous, `exp(beta)` for categorical) | Resource 2, "Covariate modeling approach" |
| CL 0.269 L/day, V2 3.02 L, Q2 1.05 L/day, V3 2.67 L | Resource 10, "Typical parameters" |
| Absorption half-life KAThalf 3.02 day (`lka = log(log(2)/3.02)`) | Resource 10 |
| Fa1 0.539; Fa1S220 0.457; Fa1S30 0.688 | Resource 10 |
| Rfa1Dose -0.554 (-42.5%); RCLADA 0.762 (+114%) | Resource 10, footnote a |
| Beta_CL BWGT 0.849; Beta_V2 BWGT 0.799; Beta_V3 BWGT 0.639; centered at 70 kg | Resource 10 and its footnote |
| Beta_CL ASIAN_1 0.0952 (+9.99%) | Resource 10 |
| All popPK omegas (reported as **standard deviations**) | Resource 10, footnote c |
| Residual errors ADD1 0.175, ADD2 0.549, ADD3 0.367 log(ng/mL) | Resource 10 and its footnote |
| Terminal half-life 15.5 day | Results 3.2 |
| AAER model equation (integral of an exponential rate with an Emax term) | Resource 3, "Longitudinal asthma exacerbation rate model" |
| AAER Base -6.64, Emax -0.51, EC50 1.76 ng/mL, omega(Base) 1.04 | Resource 17 |
| AAER beta_Base PE 0.17, OCS 0.345, EE -0.329; beta_Emax CHINESE 1.27 | Resource 17 |
| FEV1 model equation | Resource 4 |
| FEV1 Emax 104 mL, T50 9.21 day, PmaxB 172 mL, FEVB 1710 mL, KPBL 0.0356 /day | Resource 4, "base/final model" column |
| FEV1 betas (SEXF -0.09, CTHEO -0.089, BEOSL 0.699 at 380 cells/uL) | Resource 4 and its footnote |
| FEV1 residual error 137 mL additive + 0.0883 proportional | Resource 4 |
| Simulated steady-state exposures used as validation targets | Resource 12 (post-hoc) and Resource 14 (IIV-sampled) |
| Steady-state AAER ratios | Results 3.4.2 and Figure 6 |

# Closed-form structural gates

Two checks depend on no simulation at all and go red on any mis-transcribed
disposition parameter. They are the cheapest protection against a transcription
error and are run before anything stochastic.

## Terminal half-life

The paper reports a model-predicted terminal half-life of 15.5 days. For a
two-compartment model the terminal rate constant is the smaller eigenvalue of
the micro-constant system, so this is computable in closed form from CL, V2, Q2
and V3 alone. Evaluated at the reference covariates (70 kg, non-Asian):

```{r half-life}
cl_ref <- 0.269; vc_ref <- 3.02; q_ref <- 1.05; vp_ref <- 2.67

k10 <- cl_ref / vc_ref
k12 <- q_ref / vc_ref
k21 <- q_ref / vp_ref

# Smaller root of lambda^2 - (k10 + k12 + k21) lambda + k10 k21 = 0
ksum  <- k10 + k12 + k21
kprod <- k10 * k21
beta_rate <- (ksum - sqrt(ksum^2 - 4 * kprod)) / 2
thalf_terminal <- log(2) / beta_rate

c(published_day = 15.5, computed_day = round(thalf_terminal, 2))

# Deterministic quantity, so a tight bound is correct here: any of the four
# disposition parameters being wrong moves this by well over 0.2 day.
stopifnot(abs(thalf_terminal - 15.5) < 0.2)
```

## Average steady-state concentration is fixed by mass balance

At steady state on a fixed interval, the average concentration must satisfy
`Cavg = Dose * F / (CL * tau)` exactly, independent of the distribution
parameters. For a typical Chinese adult at the published mean body weight of
64.1 kg on 30 mg every 8 weeks:

```{r cavg-identity}
wt_chinese_adult <- 64.1
cl_chinese <- cl_ref * (wt_chinese_adult / 70)^0.849 * exp(0.0952)
tau <- 56  # days; every 8 weeks
fa1 <- 0.539

cavg_closed_form <- 30 * fa1 / (cl_chinese * tau) * 1000  # ng/mL

c(published_median_ngml = 1067,
  closed_form_ngml = round(cavg_closed_form, 1))

# Resource 14 reports the median of the simulated Chinese adult Cavg; the
# median of a cohort whose IIV is lognormal on CL sits close to, but not
# exactly at, the typical-value prediction.
stopifnot(abs(cavg_closed_form - 1067) / 1067 < 0.05)
```

## Weight bands implied by the published paediatric exposures

Resource 14 reports simulated steady-state exposures for Chinese adults,
adolescents, children weighing less than 35 kg (on 10 mg every 8 weeks) and
children weighing at least 35 kg (on 30 mg every 8 weeks). The adolescent and
paediatric body weights were sampled from the Chinese growth references of Zong
and Li (2013) via the R package `childsds`, which is not a dependency of this
package, so those weight distributions cannot be reproduced here directly.

The mass-balance identity above can be inverted instead: given a published
median Cavg and the dose actually used in that band, the implied median body
weight follows from the weight-on-clearance relationship. This validates the
weight-clearance chain **and** the paper's weight-based dose-selection rule
without needing the growth chart, and it is a stronger check than it looks,
because the four bands must come out monotone and physiologically plausible.

```{r implied-weights}
implied_weight <- function(cavg_ngml, dose_mg) {
  cl_needed <- dose_mg * fa1 / (cavg_ngml / 1000 * tau)
  70 * (cl_needed / (cl_ref * exp(0.0952)))^(1 / 0.849)
}

bands <- tibble::tibble(
  Band = c("Children < 35 kg", "Children >= 35 kg", "Adolescents", "Adults"),
  `Dose (mg Q8W)` = c(10, 30, 30, 30),
  `Published median Cavg (ng/mL)` = c(757.3, 1525, 1279, 1067)
) |>
  mutate(`Implied median weight (kg)` =
           round(implied_weight(`Published median Cavg (ng/mL)`, `Dose (mg Q8W)`), 1))

knitr::kable(bands, align = c("l", "r", "r", "r"),
             caption = "Median body weight implied by each band's published steady-state Cavg (Jin 2025 Resource 14).")
```

The implied weights are ordered as the bands are, and each is plausible for its
age range: roughly 26 kg for children under 35 kg (aged 6 to under 12), 41 kg
for children at or above 35 kg, 51 kg for adolescents aged 12 to under 18, and
64 kg for adults -- the last of which reproduces the observed Chinese adult mean
of 64.1 kg in Jin 2025 Table 2 almost exactly.

```{r implied-weight-gates}
iw <- setNames(bands$`Implied median weight (kg)`, bands$Band)

stopifnot(
  # Ordering across bands is a structural consequence of the dose rule plus the
  # weight-on-CL exponent, not a noisy simulated comparison.
  iw[["Children < 35 kg"]] < 35,
  iw[["Children >= 35 kg"]] >= 35,
  iw[["Children >= 35 kg"]] < iw[["Adolescents"]],
  iw[["Adolescents"]] < iw[["Adults"]],
  # The adult band has an independently published weight to hit.
  abs(iw[["Adults"]] - 64.1) < 3
)
```

Note that the two paediatric bands straddle 35 kg in the right direction: the
sub-35 kg band implies a weight below 35 kg and the other implies a weight above
it. That is the arithmetic signature of the paper's dosing rule (10 mg below
35 kg, 30 mg at or above it) being transcribed correctly; swapping the two doses
would invert it.

# Virtual cohort

Two adult arms are simulated, one Chinese and one non-Asian, using the body
weight distributions the paper reports. Chinese adolescents and children are
covered by the closed-form band check above rather than by simulation, because
their published weight distributions are not reproducible without `childsds`.

```{r cohort}
n_per_arm <- 150L  # 150 per arm; well inside the 200-per-arm cap

# Truncated-normal weight sampling from the published mean, SD and range.
sample_weight <- function(n, mean_wt, sd_wt, lo, hi) {
  out <- numeric(0)
  while (length(out) < n) {
    draw <- rnorm(2 * n, mean_wt, sd_wt)
    out <- c(out, draw[draw >= lo & draw <= hi])
  }
  out[seq_len(n)]
}

set.seed(20250623)  # seeds the covariate draws only (see note below)

cohort <- bind_rows(
  tibble::tibble(
    arm = "Chinese adults",
    # Jin 2025 Table 2, Chinese column: 64.1 +/- 11.7 kg (40.3-112).
    WT = sample_weight(n_per_arm, 64.1, 11.7, 40.3, 112),
    RACE_ASIAN = 1
  ),
  tibble::tibble(
    arm = "Non-Asian adults",
    # Resource 6 gives a non-Asian mean of 80.4 kg; Table 2's non-Chinese SD
    # (19.2 kg) and range (40-204 kg) are used, as no non-Asian SD is printed.
    WT = sample_weight(n_per_arm, 80.4, 19.2, 40, 204),
    RACE_ASIAN = 0
  )
) |>
  mutate(
    id = row_number(),
    ADA_POS = 0,        # reference subject is ADA-negative
    STUDY_MICP220 = 0,  # neither study stratum applies to a nominal regimen
    STUDY_AMES = 0,
    DOSE_HIGH = 0       # 30 mg is not the 200 mg high-dose cohort
  )

cohort |>
  group_by(arm) |>
  summarise(n = n(), `Mean WT (kg)` = round(mean(WT), 1),
            `Median WT (kg)` = round(median(WT), 1), .groups = "drop") |>
  knitr::kable(caption = "Simulated adult cohorts.")
```

Each stochastic block below calls `rxode2::rxSetSeed()` immediately before it
so that the block is reproducible on its own rather than depending on where an
earlier block left the stream.

# Steady-state simulation

The paper's exposure simulations use a nominal regimen of benralizumab 30 mg
subcutaneously every 8 weeks, summarised at steady state (Resource 14 uses
cycle 20). Twenty Q8W doses are given and only the final dosing interval is
observed.

Observation rows carry `cmt = "central"` -- the ODE state -- together with
`dvid = 1L`, which selects the `Cc` endpoint. This model has three endpoints
(`Cc`, `Cc_early`, `Cc_micp220`) that share one prediction and differ only in
which study stratum's residual error applies, so the `dvid` is what disambiguates
them. `useLinCmt = FALSE` is required: rxode2's automatic ODE-to-`linCmt()`
conversion corrupts the `dvid`-to-`cmt` mapping for multi-endpoint models.

```{r simulate-pk}
n_cycles <- 20L
dose_times <- seq(0, by = tau, length.out = n_cycles)
last_dose <- max(dose_times)
# The absorption-rate IIV is large (omega 0.707 on the log scale), so a fast
# absorber can peak within 2-3 days with sharp curvature. A uniform 1-day grid
# under-resolves that peak and understates AUC by enough to break the mass
# balance check below, so the first three weeks are sampled every 6 h.
obs_times <- last_dose + sort(unique(c(seq(0, 21, by = 0.25), seq(21, tau, by = 1))))

doses <- cohort |>
  tidyr::expand_grid(time = dose_times) |>
  mutate(amt = 30, evid = 1L, cmt = "depot", dvid = NA_integer_)

obs <- cohort |>
  tidyr::expand_grid(time = obs_times) |>
  mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L)

events <- bind_rows(doses, obs) |> arrange(id, time, desc(evid))

rxode2::rxSetSeed(20250623)
sim_pk <- rxode2::rxSolve(
  mod_pk, events,
  keep = c("arm", "WT", "RACE_ASIAN"),
  useLinCmt = FALSE,
  returnType = "data.frame"
)

# rxSolve returns one row per OBSERVATION record and no `evid` column, so the
# dose rows are already absent and no filtering is needed here.
sim_pk <- sim_pk |>
  mutate(Cc_ngml = Cc * 1000)  # model carries mg/L; the paper reports ng/mL

stopifnot(nrow(sim_pk) > 0, all(sim_pk$Cc_ngml >= 0))
range(sim_pk$time)
```

```{r plot-profiles}
sim_pk |>
  mutate(day_in_interval = time - last_dose) |>
  group_by(arm, day_in_interval) |>
  summarise(
    median = median(Cc_ngml),
    p05 = quantile(Cc_ngml, 0.05),
    p95 = quantile(Cc_ngml, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(day_in_interval, median, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.18, colour = NA) +
  geom_line(linewidth = 0.9) +
  labs(
    x = "Day within the steady-state dosing interval",
    y = "Serum benralizumab (ng/mL)",
    colour = NULL, fill = NULL,
    title = "Simulated steady-state benralizumab, 30 mg SC Q8W",
    subtitle = "Median with 5th-95th percentile band; replicates the exposure basis of Jin 2025 Figure 4"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

# PKNCA validation

Non-compartmental analysis over the final dosing interval gives Cmax, Tmax,
Cavg and the trough.

The trough is taken as PKNCA's `cmin`, the minimum concentration within the
interval. Two points make that the right choice here. First, the interval runs
from one dose to the next, so the profile goes trough, peak, trough and its
minimum is the trough by construction; at steady state the concentrations at
the two ends are equal, which the typical-value section below verifies
numerically. Second, PKNCA's `ctrough` is defined relative to a dose at the
interval end, and **no dose is given at the interval end here**, so `ctrough`
returns `NA` for every subject.

```{r pknca}
sim_nca <- sim_pk |>
  filter(!is.na(Cc_ngml)) |>
  select(id, time, Cc = Cc_ngml, arm)

# Defensive: guarantee a record at the interval start so PKNCA never
# back-extrapolates. The dose sits in the depot, so the concentration at the
# dose time is the pre-dose trough and is already in the grid; this keeps the
# guarantee explicit.
stopifnot(all(table(sim_nca$id[sim_nca$time == last_dose]) == 1))

dose_nca <- doses |> select(id, time, amt, arm)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id,
                             concu = "ng/mL", timeu = "day")
dose_obj <- PKNCA::PKNCAdose(dose_nca, amt ~ time | arm + id, doseu = "mg")

intervals <- data.frame(
  start = last_dose,
  end = last_dose + tau,
  cmax = TRUE,
  tmax = TRUE,
  cav = TRUE,
  cmin = TRUE,
  auclast = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)
```

## Mass balance of the NCA result

`AUC(0,tau)` at steady state times clearance must equal the absorbed dose.
This is an exact identity for the solved model, so it checks the numerical
accuracy of the observation grid rather than any published number, and a tight
bound is appropriate.

```{r mass-balance}
auc_by_id <- as.data.frame(nca_res$result) |>
  filter(PPTESTCD == "auclast") |>
  select(id, auclast = PPORRES)

mb <- sim_pk |>
  group_by(id, arm) |>
  summarise(cl = first(cl), fdepot = first(fdepot), .groups = "drop") |>
  mutate(id = as.character(id)) |>
  inner_join(auc_by_id |> mutate(id = as.character(id)), by = "id") |>
  mutate(
    absorbed_mg = 30 * fdepot,
    auc_x_cl_mg = auclast / 1000 * cl,  # ng/mL -> mg/L, times L/day
    pct_diff = 100 * (auc_x_cl_mg - absorbed_mg) / absorbed_mg
  )

stopifnot(nrow(mb) == 2 * n_per_arm)
summary(mb$pct_diff)

# Pure numerical (trapezoidal) error on a 1-day grid; not a cohort-noise
# quantity, so the bound can be tight.
stopifnot(max(abs(mb$pct_diff)) < 2)
```

## Typical-value steady-state exposure

Jin 2025 reports simulated steady-state exposure metrics twice. Resource 14
summarises a Chinese cohort drawn from the final model's inter-individual
variability distribution; Resource 12 summarises the *studied* adults using
post-hoc (empirical Bayes) individual parameters, so its rows reflect that
population's actual covariates rather than a resampled cohort.

Both report **medians**, and the published median is compared here against the
**typical-value** prediction rather than against the cohort median above. The
reason is reproducibility rather than convenience: a cohort median is one draw,
and `rxSetSeed()` fixes rxode2's stream only for a given solver-thread count, so
a cohort-median bound tight enough to be meaningful would pass here and fail on
a machine with a different thread count. The typical-value prediction is
deterministic. It is also the right comparator in principle, because each of
this model's random effects is lognormal and centred, so the population median
of a monotone one-eta quantity such as Cavg is the typical value.

```{r typical-value}
typical <- tibble::tibble(
  arm = c("Chinese adults", "Non-Asian adults"),
  # Jin 2025 Table 2 (Chinese mean 64.1 kg) and Resource 6 (non-Asian mean
  # 80.4 kg).
  WT = c(64.1, 80.4),
  RACE_ASIAN = c(1, 0)
) |>
  mutate(id = row_number(), ADA_POS = 0, STUDY_MICP220 = 0,
         STUDY_AMES = 0, DOSE_HIGH = 0)

ev_typ <- bind_rows(
  typical |> tidyr::expand_grid(time = dose_times) |>
    mutate(amt = 30, evid = 1L, cmt = "depot", dvid = NA_integer_),
  typical |> tidyr::expand_grid(time = obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L)
) |>
  arrange(id, time, desc(evid))

sim_typ <- rxode2::rxSolve(mod_pk, ev_typ, omega = NA,
                           keep = c("arm", "WT"), useLinCmt = FALSE,
                           returnType = "data.frame") |>
  mutate(Cc_ngml = Cc * 1000)

# Steady state is exact when the concentration at the two ends of a
# dose-to-dose interval agree. This also confirms that `cmin` over the interval
# is the trough rather than an interior minimum.
ss_check <- sim_typ |>
  group_by(arm) |>
  summarise(start_ngml = Cc_ngml[which.min(time)],
            end_ngml = Cc_ngml[which.max(time)], .groups = "drop") |>
  mutate(pct_diff = 100 * (end_ngml - start_ngml) / start_ngml)

knitr::kable(ss_check |> mutate(across(where(is.numeric), \(x) round(x, 3))),
             align = c("l", "r", "r", "r"),
             caption = "Steady state is exact: the interval's two endpoint concentrations agree.")

stopifnot(max(abs(ss_check$pct_diff)) < 0.01)
```

```{r typical-nca}
nca_typ <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(sim_typ |> select(id, time, Cc = Cc_ngml, arm),
                   Cc ~ time | arm + id, concu = "ng/mL", timeu = "day"),
  PKNCA::PKNCAdose(
    typical |> tidyr::expand_grid(time = dose_times) |>
      mutate(amt = 30) |> select(id, time, amt, arm),
    amt ~ time | arm + id, doseu = "mg"),
  intervals = intervals
))

published <- tibble::tribble(
  ~arm,                ~cmax,  ~cav,   ~cmin,
  "Chinese adults",    2214,   1067,   276.7,   # Resource 14, median column
  "Non-Asian adults",  1951,   1006,   279.8    # Resource 12, median column
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_typ,
  reference = published,
  by = "arm",
  units = c(cmax = "ng/mL", cav = "ng/mL", cmin = "ng/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Typical-value versus published steady-state benralizumab exposure (30 mg SC Q8W). * differs from the reference by more than 20%.",
  align = c("l", "l", "r", "r", "r")
)
```

```{r comparison-gates}
# `arm` is a grouping variable of the PKNCA formula, so it is already a column
# of the result table -- no join back to the cohort is needed (and a join on
# `id` would risk the type mismatch and fan-out that pknca-recipes.md warns of).
tv_val <- as.data.frame(nca_typ$result) |>
  filter(PPTESTCD %in% c("cmax", "cav", "cmin")) |>
  select(arm, PPTESTCD, PPORRES)

coh_med <- as.data.frame(nca_res$result) |>
  filter(PPTESTCD %in% c("cmax", "cav", "cmin")) |>
  group_by(arm, PPTESTCD) |>
  summarise(cohort_median = median(PPORRES), .groups = "drop")

stopifnot(nrow(tv_val) == 6L, nrow(coh_med) == 6L)

dev_tab <- published |>
  tidyr::pivot_longer(-arm, names_to = "PPTESTCD", values_to = "reference") |>
  inner_join(tv_val, by = c("arm", "PPTESTCD")) |>
  inner_join(coh_med, by = c("arm", "PPTESTCD")) |>
  mutate(
    typical = PPORRES,
    pct_diff_typical = 100 * (typical - reference) / reference,
    pct_diff_cohort = 100 * (cohort_median - reference) / reference
  ) |>
  select(arm, PPTESTCD, reference, typical, cohort_median,
         pct_diff_typical, pct_diff_cohort)

stopifnot(nrow(dev_tab) == 6L)

knitr::kable(
  dev_tab |>
    mutate(across(where(is.numeric), \(x) round(x, 1))) |>
    dplyr::rename("Arm" = arm, "Metric" = PPTESTCD,
                  "Published (ng/mL)" = reference,
                  "Typical value (ng/mL)" = typical,
                  "Cohort median (ng/mL)" = cohort_median,
                  "% diff (typical)" = pct_diff_typical,
                  "% diff (cohort)" = pct_diff_cohort),
  align = c("l", "l", "r", "r", "r", "r", "r"),
  caption = "Published medians against both the deterministic typical value and the 150-per-arm cohort median."
)

# The typical-value column is deterministic, so these bounds do not depend on
# the thread count or the draw. Realised deviations: Cmax -3.1% / -2.5%,
# Cavg -1.4% / -5.2%, Cmin -12.8% / -13.0%. A mis-transcribed clearance,
# volume, bioavailability or dose moves these by tens of percent, so the bounds
# below still go red on any transcription error.
stopifnot(
  # Cmax and Cavg are governed by CL, F and dose and are the tightest.
  max(abs(dev_tab$pct_diff_typical[dev_tab$PPTESTCD %in% c("cmax", "cav")])) < 12,
  # The trough sits on the terminal tail and depends jointly on six random
  # effects, so its published COHORT median exceeds the typical value even
  # though each individual eta is centred; see the note below.
  max(abs(dev_tab$pct_diff_typical[dev_tab$PPTESTCD == "cmin"])) < 20,
  # The cohort medians are one draw, so they are reported but bounded loosely.
  max(abs(dev_tab$pct_diff_cohort)) < 30
)
```

Cmax and Cavg reproduce to within about 5%. The trough comes out roughly 13%
below the published median in **both** arms, and the sign and near-equality of
that offset are informative: the trough lies on the terminal tail and is a
jointly non-linear function of six random effects (on CL, V2, V3, Q2, ka and
Fa1), so the cohort median is displaced upward from the typical value even
though every individual eta is centred. The 150-per-arm cohort median in the
table above bears this out, landing within about 1% of the published trough for
the Chinese arm. The offset is a property of comparing a multivariate cohort
median against a typical value, not a transcription discrepancy.

The paper's own summary of the racial difference is that Chinese adults show a
slightly higher Cmax (less than 14%) and Cavg (less than 9%) than non-Asian
adults, with no difference in trough. The typical-value predictions reproduce
that comparison:

```{r race-difference}
race_delta <- dev_tab |>
  select(arm, PPTESTCD, typical) |>
  tidyr::pivot_wider(names_from = arm, values_from = typical) |>
  mutate(
    `Typical % higher in Chinese` =
      round(100 * (`Chinese adults` - `Non-Asian adults`) / `Non-Asian adults`, 1),
    # From the Resource 12 medians: 2211 vs 1951, 1088 vs 1006, 259.1 vs 279.8.
    `Published % higher in Chinese` = c(13.3, 8.2, -7.4)
  )

knitr::kable(
  race_delta |>
    mutate(across(c(`Chinese adults`, `Non-Asian adults`), \(x) round(x, 1))) |>
    dplyr::rename("Metric" = PPTESTCD),
  align = c("l", "r", "r", "r", "r"),
  caption = "Chinese versus non-Asian typical-value steady-state exposure."
)

# The claim under test is that the racial difference is SMALL. Assert on
# magnitude, not on sign: the trough difference is near zero, so its sign is
# not a meaningful quantity (failure pattern 12).
stopifnot(
  all(abs(race_delta$`Typical % higher in Chinese`) < 25),
  # Cmax and Cavg must both be higher in Chinese participants, which IS a
  # robust ordering here: it follows deterministically from the lower Chinese
  # body weight plus the +9.99% Asian effect on clearance.
  race_delta$`Typical % higher in Chinese`[race_delta$PPTESTCD == "cmax"] > 0,
  race_delta$`Typical % higher in Chinese`[race_delta$PPTESTCD == "cav"] > 0
)
```

# Exposure-response: asthma exacerbation rate

Resource 3 writes the expected number of exacerbations in an interval as the
time integral of an instantaneous rate,

```
lambda_j(a, b) = integral_a^b exp(Base + beta * X_j + Emax_j * C(t) / (C(t) + EC50) + eta_j) dt
```

so `Jin_2025_benralizumab_aaer` carries that integral as the state
`cumhaz_exac` and reads interval counts off its increments.

## The Chinese Emax covariate form

The Chinese covariate scales Emax **fractionally**, `Emax * (1 + 1.27)`, not on
a log scale. This is settled by the paper's own headline number: the main text
says the maximal treatment effect "significantly increased (+127%)" in Chinese
participants, and 1.27 is exactly +127% under the fractional form. The two
alternatives are excluded arithmetically.

```{r emax-form}
emax_base <- -0.51
beta_chinese <- 1.27

# Emax is negative (it reduces the log rate), so "the maximal treatment effect
# increased by 127%" is a statement about its MAGNITUDE.
pct_magnitude <- function(e) 100 * (abs(e) - abs(emax_base)) / abs(emax_base)

forms <- tibble::tibble(
  Form = c("Fractional: Emax * (1 + beta)",
           "Log-scale: Emax * exp(beta)",
           "Additive: Emax + beta"),
  `Chinese Emax` = c(emax_base * (1 + beta_chinese),
                     emax_base * exp(beta_chinese),
                     emax_base + beta_chinese)
) |>
  mutate(`% larger in magnitude` = round(pct_magnitude(`Chinese Emax`), 0),
         `Plateau rate ratio` = round(exp(`Chinese Emax`), 3),
         `Chinese Emax` = round(`Chinese Emax`, 3))

knitr::kable(forms, align = c("l", "r", "r", "r"),
             caption = "Only the fractional form reproduces the published 127% increase in the magnitude of Emax.")

stopifnot(
  # The fractional form reproduces the published +127% exactly: the coefficient
  # 1.27 IS the fractional increase.
  abs(pct_magnitude(emax_base * (1 + beta_chinese)) - 127) < 0.5,
  # The log-scale form would be about +256%, nowhere near the published value.
  pct_magnitude(emax_base * exp(beta_chinese)) > 250,
  # The additive form flips the sign of the treatment effect, giving a plateau
  # rate ratio ABOVE 1, i.e. benralizumab increasing exacerbations. Decisive.
  exp(emax_base + beta_chinese) > 1
)
```

## Simulated annualised rate ratios

The rate ratio versus placebo is simulated by solving the same cohort twice,
once with 30 mg Q8W dosing and once with no dosing, and taking the ratio of the
expected exacerbations accumulated over the final year. The seed is reset before
each arm so the two arms share random numbers and the ratio is not contaminated
by different eta draws.

```{r aaer-sim}
aaer_cohort <- bind_rows(
  tibble::tibble(arm = "Chinese", WT = 64.1, RACE_CHINESE = 1, RACE_ASIAN = 1),
  tibble::tibble(arm = "Non-Chinese", WT = 78.6, RACE_CHINESE = 0, RACE_ASIAN = 0)
) |>
  tidyr::expand_grid(rep = seq_len(100L)) |>
  mutate(
    id = row_number(),
    ADA_POS = 0,
    # SIROCCO / CALIMA / MIRACLE required at least two exacerbations in the
    # prior year; 3 is used as a representative value. Base is the log rate at
    # NEXAC12M = 0, which is outside the enrolled range, so this covariate must
    # be supplied for the baseline rate to be meaningful.
    NEXAC12M = 3,
    CONMED_STEROID = 0,
    REGION_EASTEUROPE = 0
  )

# One year of observation after 20 Q8W cycles, i.e. at steady state.
year_start <- last_dose - 364
aaer_obs_times <- c(year_start, last_dose)

solve_aaer <- function(with_drug) {
  ev_obs <- aaer_cohort |>
    tidyr::expand_grid(time = aaer_obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L)
  ev <- if (with_drug) {
    ev_dose <- aaer_cohort |>
      tidyr::expand_grid(time = dose_times) |>
      mutate(amt = 30, evid = 1L, cmt = "depot", dvid = NA_integer_)
    bind_rows(ev_dose, ev_obs)
  } else {
    ev_obs
  }
  ev <- ev |> arrange(id, time, desc(evid))
  rxode2::rxSetSeed(11223344)  # same stream for both arms
  rxode2::rxSolve(mod_aaer, ev, keep = c("arm", "RACE_CHINESE"),
                  useLinCmt = FALSE, returnType = "data.frame")
}

annual_exac <- function(df, label) {
  df |>
    group_by(id, arm) |>
    arrange(time, .by_group = TRUE) |>
    summarise(events = last(cumhaz_exac) - first(cumhaz_exac), .groups = "drop") |>
    mutate(regimen = label)
}

aaer_drug <- annual_exac(solve_aaer(TRUE), "Benralizumab 30 mg Q8W")
aaer_pbo  <- annual_exac(solve_aaer(FALSE), "Placebo")

aaer_ratio <- bind_rows(aaer_drug, aaer_pbo) |>
  tidyr::pivot_wider(names_from = regimen, values_from = events) |>
  group_by(arm) |>
  summarise(
    `Placebo AAER` = mean(Placebo),
    `Benralizumab AAER` = mean(`Benralizumab 30 mg Q8W`),
    `Simulated rate ratio` = mean(`Benralizumab 30 mg Q8W`) / mean(Placebo),
    .groups = "drop"
  ) |>
  mutate(
    `Published rate ratio` = c(0.38, 0.64),
    `Published 95% CI` = c("0.32-0.45", "0.60-0.71")
  )

knitr::kable(
  aaer_ratio |> mutate(across(where(is.numeric), \(x) round(x, 3))),
  align = c("l", "r", "r", "r", "r", "l"),
  caption = "Simulated versus published steady-state annual asthma exacerbation rate ratios (Jin 2025 Figure 6, adult population)."
)
```

```{r aaer-gates}
rr <- setNames(aaer_ratio$`Simulated rate ratio`, aaer_ratio$arm)

# At the exposures this model is declared valid over, EC50 (1.76 ng/mL) is
# ~150-fold below the trough, so C/(C + EC50) is essentially 1 and the rate
# ratio collapses to exp(Emax). That analytic limit is the primary gate.
plateau_nonchinese <- exp(emax_base)
plateau_chinese <- exp(emax_base * (1 + beta_chinese))

c(simulated_nonchinese = round(rr[["Non-Chinese"]], 3),
  plateau_nonchinese = round(plateau_nonchinese, 3),
  simulated_chinese = round(rr[["Chinese"]], 3),
  plateau_chinese = round(plateau_chinese, 3))

stopifnot(
  # The simulation must land on the analytic plateau, which it does to within a
  # fraction of a percent; this is a deterministic property of the model.
  abs(rr[["Non-Chinese"]] - plateau_nonchinese) / plateau_nonchinese < 0.02,
  abs(rr[["Chinese"]] - plateau_chinese) / plateau_chinese < 0.02,
  # The Chinese arm must show the substantially larger effect.
  rr[["Chinese"]] < rr[["Non-Chinese"]],
  # Both must sit in the neighbourhood of the published ratios. The Chinese arm
  # is a documented deviation (see Assumptions and deviations) and is bounded
  # more loosely; 0.30 still goes red on a mis-signed or mis-scaled Emax.
  abs(rr[["Non-Chinese"]] - 0.64) < 0.10,
  abs(rr[["Chinese"]] - 0.38) < 0.12
)
```

The non-Chinese ratio reproduces the published 0.64 well and falls inside the
published 95% CI of 0.60 to 0.71. The Chinese ratio comes out lower than the
published 0.38; see the deviations section below.

# Exposure-response: pre-bronchodilator FEV1

The FEV1 model is algebraic and exposure-independent. Its three terms separate
cleanly, so each can be checked against the published parameter it comes from.

```{r fev1-typical}
fev1_grid <- tibble::tibble(time = seq(0, 365, by = 1))

fev1_arms <- bind_rows(
  tibble::tibble(arm = "Placebo", TRT_BENRALIZUMAB = 0),
  tibble::tibble(arm = "Benralizumab", TRT_BENRALIZUMAB = 1)
) |>
  mutate(
    id = row_number(),
    SEXF = 0,                 # male reference
    CONMED_THEOPHYLLINE = 0,  # non-user reference
    EOS = 380                 # the reference eosinophil count of the covariate
  )

ev_fev1 <- fev1_arms |>
  tidyr::expand_grid(fev1_grid) |>
  mutate(amt = NA_real_, evid = 0L)

sim_fev1 <- rxode2::rxSolve(mod_fev1, ev_fev1, omega = NA,
                            keep = c("arm", "TRT_BENRALIZUMAB"),
                            returnType = "data.frame")

sim_fev1 |>
  ggplot(aes(time, FEV1, colour = arm)) +
  geom_line(linewidth = 0.9) +
  labs(x = "Time (day)", y = "Pre-bronchodilator FEV1 (mL)", colour = NULL,
       title = "Typical-value longitudinal FEV1, reference subject",
       subtitle = "Male, no theophylline, baseline eosinophils 380 cells/uL (Jin 2025 Resource 4)") +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r fev1-decomposition}
fev1_at <- function(a, tt) {
  v <- sim_fev1$FEV1[sim_fev1$arm == a & sim_fev1$time == tt]
  if (length(v) != 1L) stop("no unique FEV1 for ", a, " at t = ", tt)
  v
}

decomp <- tibble::tibble(
  Quantity = c("Baseline FEV1 at t = 0",
               "Placebo plateau minus baseline at 1 year",
               "Benralizumab minus placebo at 1 year",
               "Time to half the benralizumab effect"),
  Published = c(1710, 172, 104, 9.21),
  Realised = c(
    fev1_at("Placebo", 0),
    fev1_at("Placebo", 365) - 1710,
    fev1_at("Benralizumab", 365) - fev1_at("Placebo", 365),
    # T50 recovered from the Emax term: the day the drug increment reaches half
    # of Emax. Solve emax * t / (T50 + t) = emax / 2  =>  t = T50.
    {
      drug <- sim_fev1$FEV1[sim_fev1$arm == "Benralizumab"] -
        sim_fev1$FEV1[sim_fev1$arm == "Placebo"]
      tt <- sim_fev1$time[sim_fev1$arm == "Benralizumab"]
      approx(drug, tt, xout = 104 / 2)$y
    }
  )
) |>
  mutate(`% difference` = round(100 * (Realised - Published) / Published, 2),
         Realised = round(Realised, 2))

knitr::kable(decomp, align = c("l", "r", "r", "r"),
             caption = "The FEV1 model's terms recovered from the solved typical-value profile.")

# Deterministic (typical-value, no IIV) quantities, so tight bounds are
# correct. The drug increment at 1 year is 97.5% of Emax, not 100%, because
# t / (T50 + t) = 365 / 374.21.
stopifnot(
  abs(fev1_at("Placebo", 0) - 1710) < 1,
  abs((fev1_at("Placebo", 365) - 1710) - 172) < 1,
  abs((fev1_at("Benralizumab", 365) - fev1_at("Placebo", 365)) - 104 * 365 / (9.21 + 365)) < 1,
  abs(decomp$Realised[4] - 9.21) < 0.5
)
```

The eosinophil covariate on Emax is the one covariate the FEV1 model retains on
the treatment effect, and it is the mechanistically expected one for an
anti-IL-5-receptor antibody: patients with higher baseline eosinophil counts
gain more lung function.

```{r fev1-eos}
eos_levels <- c(150, 380, 800, 1500)

eos_cohort <- tibble::tibble(EOS = eos_levels) |>
  mutate(id = row_number(), SEXF = 0, CONMED_THEOPHYLLINE = 0,
         TRT_BENRALIZUMAB = 1)

sim_eos <- rxode2::rxSolve(
  mod_fev1,
  eos_cohort |> tidyr::expand_grid(time = c(0, 365)) |>
    mutate(amt = NA_real_, evid = 0L),
  omega = NA, keep = "EOS", returnType = "data.frame"
)

eos_tab <- sim_eos |>
  group_by(EOS) |>
  summarise(fev1_365 = FEV1[time == 365], .groups = "drop") |>
  mutate(
    `Emax (mL)` = round(104 * (EOS / 380)^0.699, 1),
    `FEV1 at 1 year (mL)` = round(fev1_365, 1)
  ) |>
  select(-fev1_365) |>
  dplyr::rename("Baseline eosinophils (cells/uL)" = EOS)

knitr::kable(eos_tab, align = c("r", "r", "r"),
             caption = "Baseline eosinophil count scales the benralizumab FEV1 Emax as (EOS/380)^0.699 (Jin 2025 Resource 4).")

stopifnot(
  # Monotone increasing in eosinophils, and the reference level reproduces the
  # published Emax exactly.
  all(diff(eos_tab$`Emax (mL)`) > 0),
  abs(eos_tab$`Emax (mL)`[eos_tab$`Baseline eosinophils (cells/uL)` == 380] - 104) < 0.5
)
```

# Assumptions and deviations

## Two retained FEV1 covariate effects are omitted

Jin 2025 Resource 4 states that the FEV1 model "included covariate effects of
age and height on Pmax, age, height, sex, theophylline comedication on
FEV1baseline and eosinophil count on the Emax", and tabulates three
coefficients for the age and height effects: `RHGHT` = 2.4 ("effect of height on
PMAX and FEV1"), `RPAGE` = -5.92 ("effect of age on PMAX") and `RFAGE` = -14.6
("effect of age on FEV1").

**The height and age effects are deliberately not encoded.** Neither their
functional form nor a reference value is stated anywhere, and height is not
reported in any demographics table in Jin 2025 or its supplement, so there is no
median to centre on. The upstream legacy publication was obtained and checked
(Chia YL et al., Clin Pharmacol Ther. 2019;106:383-90, doi:10.1002/cpt.1371):
it prints the same base equation and names height and age as significant on both
Pmax and baseline FEV1, but likewise gives no functional form, no reference
value and no parameter table. Its Figures S2 and S3 are individual-prediction
diagnostic scatter plots with LOESS trends by age band, which cannot identify a
covariate form because they carry the model's (large) inter-individual
variability; Chia 2019's text further indicates the age effect on the placebo
response is not a simple linear slope, reporting that "adolescents had the
strongest placebo effect compared with other age groups (with the same standing
height)" and that "no placebo effect was observed for patients aged > 65 years".

Guessing a form and a reference height would move baseline FEV1 by hundreds of
mL, so the effects are omitted rather than invented. Both are documented in
`covariatesDataExcluded` in `Jin_2025_benralizumab_fev1`. The tabulated FEVB
(1710 mL) and PmaxB (172 mL) are consistent with being the typical values *at*
the reference covariates -- 1710 mL matches the observed mean pre-bronchodilator
FEV1 in the pivotal trials -- so omitting the centred terms leaves a coherent
typical-subject model, which is what the FEV1 section above validates.

This is a reporting gap in the literature rather than a missing file: the
relevant table is on disk, and the upstream source was acquired and does not
resolve it.

## The Chinese AAER ratio is lower than published

The simulated Chinese rate ratio (about 0.32) is below the published 0.38 and
just below its 95% CI lower bound of 0.32, while the non-Chinese ratio (about
0.60) sits inside its published CI of 0.60 to 0.71. Both simulated values land
on the analytic plateau `exp(Emax)` to within a fraction of a percent, so the
model as encoded is internally consistent; the residual gap is in how the
published ratio was computed. Jin 2025 obtained it from 200 replicates of 500
simulated participants that resample body weight, covariates and parameter
uncertainty and then form the ratio of predicted annual rates, whereas the
simulation here is a single fixed-covariate cohort with no parameter-uncertainty
resampling. The ratio of the two published logs, `log(0.38)/log(0.6419) = 2.18`,
is close to the fractional-form prediction of `1 + 1.27 = 2.27`, which is the
strongest available confirmation that the Emax covariate form is right: the
log-scale alternative predicts 3.56 and the additive alternative predicts a
rate ratio above 1. The deviation is recorded rather than tuned away.

## Discrepant omega for RCLADA between supplement tables

Resource 10 (the final-model table) prints `Omega, RCLADA` as 0.029 (fixed),
which is the same value as the `Omega, Rfa1Dose` row immediately above it.
Resource 8 and Resource 11 both print 0.036 (fixed) for RCLADA, unchanged across
every model column, and Resource 11's middle column *is* the final model --
every one of its other twenty-plus values matches Resource 10 exactly. The
0.036 value is therefore used and the 0.029 in Resource 10 is treated as a
transcription slip duplicating the row above. The parameter is a nearly
degenerate placeholder (99.3% shrinkage) so the choice has no practical effect
on any prediction, but it is recorded rather than silently resolved.

## Other assumptions

* **Absorption is reparameterised from a half-life to a rate.** Jin 2025
  estimates `KAThalf` = 3.02 days with a lognormal omega of 0.707, whereas the
  canonical parameter is `lka`. A lognormal random effect on a half-life is
  exactly a lognormal random effect of the same magnitude on the corresponding
  rate constant, with the sign of eta mirrored, which leaves the distribution
  unchanged; the omega therefore transfers unaltered.
* **The three residual errors are encoded as three endpoints.** Jin 2025 fits
  one log-scale additive residual error per study stratum. nlmixr2 selects a
  residual error by endpoint rather than by a covariate value, so the popPK
  model carries `Cc`, `Cc_early` and `Cc_micp220` over one shared prediction;
  assign each concentration record to the endpoint matching its study.
* **Steady state is simulated as a pure Q8W regimen.** The pivotal regimen gives
  the first three doses every 4 weeks and then every 8 weeks, but the paper's
  exposure simulations use a "nominal dosing regimen of 30 mg Q8W", and by
  cycle 20 (about 65 terminal half-lives) the loading detail is irrelevant.
* **The non-Asian weight SD is borrowed.** Resource 6 gives a non-Asian mean
  body weight of 80.4 kg but no SD; Table 2's non-Chinese SD of 19.2 kg and
  range of 40-204 kg are used instead.
* **Resource 12 is a post-hoc summary.** Its non-Asian row summarises the
  studied participants using empirical Bayes individual parameters, so it
  reflects that population's actual covariates rather than a resampled cohort;
  the Chinese target is taken from Resource 14, which is IIV-sampled and so is
  the like-for-like comparator for the simulation here.
* **The exposure-response N is an upper bound, not an analysis-set size.** The
  paper does not print an analysis-set N for either exposure-response dataset,
  stating only that the SIROCCO, CALIMA and MIRACLE data were used. Both ER model
  files therefore set `population$n_subjects` to the combined *randomised* total
  of those three studies (1204 + 1306 + 695 = 3205) and say so explicitly. The
  only exposure-response counts the paper enumerates are the per-quartile Q8W
  benralizumab recipients in Resource 15 (1023 in total), to which the placebo
  and 30 mg Q4W arms must be added but are not tabulated. For the FEV1 model no
  single N describes the whole fit in any case, since only benralizumab-treated
  patients contributed to estimating the treatment Emax while all arms
  contributed to the baseline and placebo terms.
* **Model validity range.** Jin 2025 states that the exacerbation-rate model
  "is valid for doses leading to mean average concentrations above 500 ng/mL and
  mean Ctrough above 50 ng/mL". EC50 (1.76 ng/mL) is below the assay LLOQ of
  3.86 ng/mL and carries a 180% RSE, so the exposure-response curve is flat and
  at its plateau across the studied dose range. Do not use either ER model to
  extrapolate to lower exposures.
* **No paediatric simulation is run.** The adolescent and paediatric weight
  distributions come from the Chinese growth references of Zong and Li (2013)
  via the `childsds` package, which is not a dependency here. The closed-form
  implied-weight check above validates the same chain without it.

# Session information

```{r session-info}
sessionInfo()
```
