---
title: "Lumefantrine (Juma 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Lumefantrine (Juma 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
mod_fn <- readModelDb("Juma_2026_lumefantrine")
mod    <- rxode2::rxode2(mod_fn())
```

* Citation: `r mod$reference`
* Description: `r mod$description`
* Article: <https://doi.org/10.1111/bcp.70318>

Juma and colleagues studied lumefantrine after the standard six-dose
artemether-lumefantrine regimen in pregnant and non-pregnant Kenyan women with
uncomplicated *Plasmodium falciparum* malaria, and found that pregnancy raises
apparent clearance by 23.2% and apparent central volume by 28.1%, lowering
lumefantrine exposure by roughly 30%.

# Population

Fifty pregnant women (30 in the second and 20 in the third trimester, gestational
age median 26 weeks, range 13-40) and 25 non-pregnant women were enrolled at
Ahero County Hospital in Kisumu County, western Kenya, between August 2013 and
April 2014 (Pan African Clinical Trial Registry PACTR201211000451437). All
participants were women aged 18-40 years with uncomplicated *P. falciparum*
mono-infection (1000-200 000 parasites/uL), haemoglobin at least 8 g/dL, and no
antimalarial exposure in the preceding month.

Baseline characteristics are in Table 1 of the source: median age 20.2 years
(18.0-35.3) in pregnant and 25.2 years (18.1-35.6) in non-pregnant women; median
body weight 62.0 kg (40.0-86.6) and 59.5 kg (45.0-89.0) respectively; median
enrolment parasitaemia 10 960/uL and 24 320/uL. Every participant received four
tablets of artemether/lumefantrine 20/120 mg (Coartem, Novartis AG), i.e. 480 mg
lumefantrine per dose, at 0, 8, 24, 36, 48 and 60 h, taken with 250 mL of milk or
a light meal. The day-28 PCR-corrected cure rate was 100% in both groups.

The population PK model was built in NONMEM 7.4 (FOCE-I) from 1151 venous plasma
lumefantrine concentrations; 86 samples (7.4%) were below the 10 ng/mL LLOQ and
were omitted (M1). Parameter uncertainty came from sampling importance
resampling.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Juma_2026_lumefantrine")()$population`).

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Juma_2026_lumefantrine.R` carries an in-file comment
naming its source location. They are collected here for review.

| Equation / parameter | Value as encoded | Source location |
|---|---|---|
| `lmtt` | `log(4.70)` h | Table 2, `MTT (h)` = 4.70 (RSE 6.9%) |
| transit chain length | 3 transit compartments | Table 2, `No. transit comp.` = 3 (fixed); Figure 1 |
| `lka` | `log(0.411)` 1/h | Table 2, `Ka (1/h)` = 0.411 (RSE 12.7%) |
| `lcl` | `log(4.03)` L/h | Table 2, `CL/F (L/h)` = 4.03 (RSE 5.1%) |
| `lvc` | `log(108)` L | Table 2, `VC/F (L)` = 108 (RSE 8.6%) |
| `lq` | `log(1.67)` L/h | Table 2, `Q/F (L/h)` = 1.67 (RSE 7.1%) |
| `lvp` | `log(184)` L | Table 2, `Vp/F (L)` = 184 (RSE 6.1%) |
| `lfdepot` | `fixed(log(1))` | Table 2, `F (%)` = 100 (fixed); Methods 2.9 |
| `boxcox_fdepot` | -0.690 | Table 2, `Box-cox on F` = -0.690 (RSE 13.2%); Results 3.1.2 |
| `e_wt_cl` | `fixed(0.75)` | Methods 2.11 and Equation 2 |
| `e_wt_vc` | `fixed(1)` | Methods 2.11 and Equation 3 |
| weight centring | 70 kg | Methods 2.11; Table 2 note |
| `dose50` | `fixed(3.84)` mg/kg | Table 2, `Dose50 (mg/kg) on F` (fixed); Methods 2.11, Equation 4 |
| `e_preg_cl` | 0.232 | Table 2, `Pregnancy on CL (%)` = 23.2; Results 3.1.2 |
| `e_preg_vc` | 0.281 | Table 2, `Pregnancy on Vc (%)` = 28.1; Results 3.1.2 |
| `etalmtt` | 0.4405336 | Table 2, `MTT (IIV)` = 74.4% CV; Table 2 note CV formula |
| `etalvc` | 0.0196860 | Table 2, `VC/F (IIV)` = 14.1% CV |
| `etaiov_ka_1`, `etaiov_ka_2` | 0.7830336 | Table 2, `Ka (IOV)` = 109% CV; Methods 2.10 (two pooled occasions) |
| `etaiov_fdepot_1`, `etaiov_fdepot_2` | 0.3433060 | Table 2, `F (IOV)` = 64.0% CV; Methods 2.10 |
| `propSd` | 0.330 | Table 2, `Proportional (%)` = 33.0 |
| `addSd` | 0.0120 ug/mL | Table 2, `Additive (ng/mL)` = 12.0 |
| absorption ODEs (`depot`, `transit1-3`) | n/a | Figure 1; Methods 2.9; Results 3.1.2 |
| two-compartment disposition ODEs | n/a | Figure 1; Results 3.1.2 |
| `f(depot)` composition | n/a | Equation 4; Table 2 note; Methods 2.9 |
| pregnancy proportional form | n/a | Table 2 note: `theta = theta_TV * (1 + theta_cov)` |
| CV-to-variance conversion | n/a | Table 2 note: `CV% = 100 * sqrt(exp(variance) - 1)` |

# Structural integrity checks

These are cheap, deterministic gates on the encoded structure. They do not
depend on any random draw, so they hold on any machine and any rxode2 build.

```{r structural}
# 1. The explicit ODE system must actually be solved. If rxode2 recognised a
#    cl/vc pair and switched to its analytic one-compartment solution, the
#    transit chain feeding `central` would be silently discarded.
stopifnot(length(mod$linCmt) == 0)
stopifnot(identical(
  mod$state,
  c("depot", "transit1", "transit2", "transit3", "central", "peripheral1")
))

dose_times <- c(0, 8, 24, 36, 48, 60)
obs_times  <- sort(unique(c(seq(0, 72, by = 0.5), seq(78, 672, by = 6))))

make_events <- function(wt, preg, cohort, id_offset = 0L) {
  n <- length(wt)
  dosing <- tidyr::expand_grid(i = seq_len(n), time = dose_times) |>
    dplyr::mutate(amt = 480, evid = 1L, cmt = "depot")
  obs <- tidyr::expand_grid(i = seq_len(n), time = obs_times) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
  dplyr::bind_rows(dosing, obs) |>
    dplyr::mutate(
      id     = id_offset + i,
      WT     = wt[i],
      PREG   = preg,
      DOSE   = 480,
      # Methods 2.10: doses 1-3 are occasion 1, doses 4-6 are occasion 2.
      OCC    = ifelse(time < 36, 1, 2),
      cohort = cohort
    ) |>
    dplyr::select(id, time, amt, evid, cmt, WT, PREG, DOSE, OCC, cohort) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

mod_typ <- rxode2::zeroRe(mod)
ev_typ  <- make_events(70, 0L, "typical")
s_typ   <- rxode2::rxSolve(mod_typ, ev_typ, returnType = "data.frame",
                           addDosing = FALSE)

# 2. Mass balance. At the end of a completed course, CL/F * AUC(0, Inf) must
#    equal the total administered dose times the bioavailability actually
#    applied. This is the one check a silently-zeroed absorption chain cannot
#    pass, and a ratio-free equality also catches a dose arriving twice.
f_dose  <- 1 - (480 / 70) / (3.84 + 480 / 70)
lam_idx <- c(nrow(s_typ) - 24L, nrow(s_typ))
lambda_z <- -diff(log(s_typ$Cc[lam_idx])) / diff(s_typ$time[lam_idx])
auc_obs  <- sum(diff(s_typ$time) *
                  (head(s_typ$Cc, -1) + tail(s_typ$Cc, -1)) / 2)
auc_inf  <- auc_obs + tail(s_typ$Cc, 1) / lambda_z
mass_balance <- 4.03 * auc_inf / (6 * 480 * f_dose)
stopifnot(abs(mass_balance - 1) < 0.005)

# 3. The absorption really is a transit chain, not a bolus into `depot`.
#    A bolus would put the whole F-adjusted dose in `depot` immediately.
depot_early <- s_typ$depot[which.min(abs(s_typ$time - 0.5))]
stopifnot(depot_early < 0.85 * 480 * f_dose, depot_early > 0)

# 4. MTT sensitivity, on a single dose so that Tmax is unambiguous.
#    Quadrupling the mean transit time must visibly delay the peak; an
#    absorption chain that had been discarded would leave Tmax unmoved.
ev_single <- ev_typ |> dplyr::filter(evid == 0 | time == 0)
tmax_at <- function(m) {
  s <- rxode2::rxSolve(m, ev_single, returnType = "data.frame",
                       addDosing = FALSE)
  s$time[which.max(s$Cc)]
}
mod_slow  <- rxode2::zeroRe(rxode2::ini(mod, lmtt = log(4 * 4.70)))
tmax_fast <- tmax_at(mod_typ)
tmax_slow <- tmax_at(mod_slow)
stopifnot(tmax_slow - tmax_fast > 5)

c(mass_balance = round(mass_balance, 6),
  terminal_half_life_h = round(log(2) / lambda_z, 1),
  f_applied = round(f_dose, 4),
  single_dose_tmax_h = tmax_fast,
  single_dose_tmax_4x_mtt_h = tmax_slow)
```

The terminal half-life of the typical-value profile is close to the 109 h
reported for non-pregnant women in Table 3, and `CL/F * AUC(0, Inf)` recovers the
administered dose times the applied bioavailability to within rounding.

# Pregnancy effect (replicates Figure 5)

The published geometric mean ratios for pregnant versus non-pregnant women come
from the final model's covariate effects, so they can be reproduced exactly with
a typical-value solve holding body weight fixed at the model's 70 kg reference.
That makes this comparison independent of any simulated cohort.

```{r figure-5}
solve_typ <- function(preg) {
  rxode2::rxSolve(
    mod_typ, make_events(70, preg, "typical"),
    returnType = "data.frame", addDosing = FALSE
  )
}
np   <- solve_typ(0L)
preg <- solve_typ(1L)

auc_of <- function(x) {
  lam <- -diff(log(x$Cc[c(nrow(x) - 24L, nrow(x))])) /
    diff(x$time[c(nrow(x) - 24L, nrow(x))])
  sum(diff(x$time) * (head(x$Cc, -1) + tail(x$Cc, -1)) / 2) +
    tail(x$Cc, 1) / lam
}
d7_of <- function(x) x$Cc[which.min(abs(x$time - 168))] * 1000

gmr <- tibble::tibble(
  Parameter = c("AUC(0, Inf)", "Cmax", "Day-7 concentration"),
  `Non-pregnant` = c(auc_of(np), max(np$Cc), d7_of(np)),
  `Pregnant` = c(auc_of(preg), max(preg$Cc), d7_of(preg)),
  `Model ratio` = `Pregnant` / `Non-pregnant`,
  `Published GMR (95% CI)` = c("0.76 (0.57-1.01)", "0.79 (0.63-0.99)",
                               "0.69 (0.51-0.94)"),
  lo = c(0.57, 0.63, 0.51),
  hi = c(1.01, 0.99, 0.94)
)

# Every model-derived ratio must fall inside the published 95% CI. This is a
# direct test of the two pregnancy coefficients and of the proportional
# covariate form, and it is deterministic.
stopifnot(all(gmr$`Model ratio` >= gmr$lo & gmr$`Model ratio` <= gmr$hi))

gmr |>
  dplyr::select(-lo, -hi) |>
  knitr::kable(
    digits = 3,
    caption = paste(
      "Replicates Figure 5 of Juma 2026: pregnancy effect on secondary PK",
      "parameters, expressed as the pregnant / non-pregnant ratio for a",
      "typical 70 kg woman. Published geometric mean ratios are quoted from",
      "the Results section."
    )
  )
```

# Virtual cohort

The original concentration data are not public. The cohorts below reproduce the
three groups of Table 3 using the body-weight distributions reported in Table 1
and in Results 3.1: non-pregnant median 59.5 kg (45.0-89.0), second-trimester
median 57 kg (40-74), third-trimester median 66 kg (51-86). Weights are drawn
log-normally around each median and truncated to the reported range. The model
has no trimester term, so the only thing separating the two pregnant arms is
body weight.

```{r cohort}
# set.seed() seeds R's RNG, which draws the covariates below. It does NOT seed
# rxode2's simulation RNG, whose streams are partitioned per solver thread, so
# the etas differ between a 2-core CI runner and a many-thread workstation.
# Every assertion downstream is therefore written on medians and robust
# quantiles, never on a cohort extreme.
set.seed(20260912)
n_arm <- 150L

draw_wt <- function(n, med, lo, hi, sdlog) {
  w <- rlnorm(n, log(med), sdlog)
  pmin(pmax(w, lo), hi)
}

events <- dplyr::bind_rows(
  make_events(draw_wt(n_arm, 59.5, 45, 89, 0.17), 0L,
              "Non-pregnant", id_offset = 0L),
  make_events(draw_wt(n_arm, 57.0, 40, 74, 0.16), 1L,
              "Pregnant, 2nd trimester", id_offset = 1000L),
  make_events(draw_wt(n_arm, 66.0, 51, 86, 0.14), 1L,
              "Pregnant, 3rd trimester", id_offset = 2000L)
)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
rxode2::rxSetSeed(20260912)
sim <- rxode2::rxSolve(mod, events = events, keep = c("cohort", "WT"),
                       addDosing = FALSE)
sim <- as.data.frame(sim)
stopifnot(!anyNA(sim$Cc), all(sim$Cc >= 0))
```

# Replicate published figures

```{r figure-4, fig.height = 5}
# Replicates Figure 4 of Juma 2026: predicted lumefantrine concentration-time
# profiles in non-pregnant (panel A) and pregnant (panel B) women. The paper's
# figure is a visual predictive check against observed data; the equivalent
# here is the simulated 5th / 50th / 95th percentile envelope.
sim |>
  dplyr::group_by(cohort, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = median(Cc),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  dplyr::filter(Q05 > 0) |>
  ggplot2::ggplot(ggplot2::aes(time / 24, Q50)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  ggplot2::geom_line() +
  ggplot2::geom_hline(yintercept = 0.200, linetype = "dashed",
                      colour = "firebrick") +
  ggplot2::facet_wrap(~cohort) +
  ggplot2::scale_y_log10() +
  ggplot2::labs(
    x = "Time (days)", y = "Lumefantrine concentration (ug/mL)",
    title = "Simulated lumefantrine profiles by pregnancy group",
    caption = paste(
      "Replicates Figure 4 of Juma 2026. Median with 5th-95th percentile",
      "envelope. Dashed line is the 200 ng/mL day-7 efficacy threshold."
    )
  )
```

```{r figure-4-early}
# The lower panels of Figure 4 zoom on the first 72 h of treatment.
sim |>
  dplyr::filter(time <= 72) |>
  dplyr::group_by(cohort, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = median(Cc),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot2::ggplot(ggplot2::aes(time, Q50)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  ggplot2::geom_line() +
  ggplot2::facet_wrap(~cohort) +
  ggplot2::labs(
    x = "Time (h)", y = "Lumefantrine concentration (ug/mL)",
    title = "First 72 h of the six-dose regimen",
    caption = "Replicates the lower panels of Figure 4 of Juma 2026."
  )
```

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, cohort)

# Guarantee a pre-dose record so PKNCA can anchor AUC at time zero; for an
# oral model the correct pre-dose concentration is zero.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, cohort) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, cohort, time, .keep_all = TRUE) |>
  dplyr::arrange(cohort, id, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | cohort + id)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, cohort)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | cohort + id)

# The paper's secondary parameters are for the whole six-dose course: Cmax is
# the peak over the regimen and AUC runs from time zero to infinity.
intervals <- data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  auclast    = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)

nca_tbl <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  dplyr::select(cohort, id, PPTESTCD, PPORRES)
stopifnot(nrow(nca_tbl) > 0, !anyNA(nca_tbl$PPORRES))

# Day-7 concentration is read directly off the simulation, not from PKNCA.
day7 <- sim |>
  dplyr::filter(abs(time - 168) < 1e-6) |>
  dplyr::transmute(cohort, id, PPTESTCD = "day7", PPORRES = Cc * 1000)

nca_all <- dplyr::bind_rows(nca_tbl, day7)
```

## Comparison against the published secondary parameters

```{r nca-compare}
simulated <- nca_all |>
  dplyr::group_by(cohort, PPTESTCD) |>
  dplyr::summarise(median = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median)

# Table 3 of Juma 2026, median (95% CI) across the Bayesian post-hoc estimates.
published <- tibble::tribble(
  ~cohort,                    ~half.life, ~cmax, ~aucinf.obs, ~day7,
  "Non-pregnant",             109,        3.21,  275,         277,
  "Pregnant, 2nd trimester",  103,        2.28,  184,         182,
  "Pregnant, 3rd trimester",  106,        2.44,  208,         192
)

cmp <- simulated |>
  dplyr::select(cohort, half.life, cmax, aucinf.obs, tmax, day7) |>
  dplyr::left_join(published, by = "cohort", suffix = c("_sim", "_pub")) |>
  dplyr::mutate(
    dplyr::across(dplyr::ends_with(c("_sim", "_pub")), ~ signif(.x, 3))
  )

cmp |>
  dplyr::select(
    cohort,
    half.life_sim, half.life_pub,
    cmax_sim, cmax_pub,
    aucinf.obs_sim, aucinf.obs_pub,
    day7_sim, day7_pub,
    tmax
  ) |>
  dplyr::rename(
    "Group"                    = cohort,
    "t1/2 sim (h)"             = half.life_sim,
    "t1/2 published (h)"       = half.life_pub,
    "Cmax sim (ug/mL)"         = cmax_sim,
    "Cmax published (ug/mL)"   = cmax_pub,
    "AUC sim (h*ug/mL)"        = aucinf.obs_sim,
    "AUC published (h*ug/mL)"  = aucinf.obs_pub,
    "Day 7 sim (ng/mL)"        = day7_sim,
    "Day 7 published (ng/mL)"  = day7_pub,
    "Tmax sim (h)"             = tmax
  ) |>
  knitr::kable(
    caption = paste(
      "Simulated versus published secondary PK parameters (Table 3 of",
      "Juma 2026). Simulated values are cohort medians; published values are",
      "medians of the Bayesian post-hoc estimates from the study cohorts.",
      "Tmax is not reported in the source."
    )
  )
```

```{r nca-assertions}
# Assert on the CENTRE of each cohort, never on its extremes: the extreme of a
# random cohort is not reproducible across rxode2 builds, whereas the median is
# driven by the structural parameters and the covariate model.
rel <- function(sim, pub) abs(sim - pub) / pub

chk <- cmp |>
  dplyr::transmute(
    cohort,
    hl   = rel(half.life_sim, half.life_pub),
    cmax = rel(cmax_sim, cmax_pub),
    auc  = rel(aucinf.obs_sim, aucinf.obs_pub),
    d7   = rel(day7_sim, day7_pub)
  )

stopifnot(
  # Terminal half-life is a pure function of the disposition parameters and is
  # essentially free of cohort noise, so it can carry a tight bound.
  all(chk$hl < 0.15),
  # Exposure medians compare a virtual cohort against post-hoc medians from the
  # real cohort, whose covariate distribution is only approximated here.
  all(chk$cmax < 0.30),
  all(chk$auc  < 0.35),
  all(chk$d7   < 0.35),
  # The non-pregnant arm is the model's reference group and its weight
  # distribution is the one reported directly in Table 1, so it is held to a
  # tighter bound than the two pregnant arms.
  chk$cmax[chk$cohort == "Non-pregnant"] < 0.20,
  chk$auc[chk$cohort == "Non-pregnant"]  < 0.20,
  chk$d7[chk$cohort == "Non-pregnant"]   < 0.20
)
round(chk[, -1] * 100, 1)
```

```{r day7-threshold}
# Results 3.1.2 and the Discussion report that 44% of pregnant and 20% of
# non-pregnant women had a day-7 concentration below the 200 ng/mL threshold
# associated with therapeutic failure. Reproduced descriptively -- the source
# percentages come from the observed cohort of 50 and 25 women, so they are
# reported here rather than asserted on.
nca_all |>
  dplyr::filter(PPTESTCD == "day7") |>
  dplyr::group_by(cohort) |>
  dplyr::summarise(
    `Median day-7 (ng/mL)` = round(median(PPORRES)),
    `Below 200 ng/mL (%)`  = round(100 * mean(PPORRES < 200), 1),
    .groups = "drop"
  ) |>
  dplyr::rename("Group" = cohort) |>
  knitr::kable(
    caption = paste(
      "Day-7 lumefantrine concentrations against the 200 ng/mL efficacy",
      "threshold. Juma 2026 reports 44% of pregnant and 20% of non-pregnant",
      "women below this threshold."
    )
  )
```

# Assumptions and deviations

* **Allometric centring weight.** Methods 2.11 states the allometric function
  "was centred on a typical body weight of 70 kg", and the Table 2 note confirms
  that the population estimates are for a typical non-pregnant woman weighing
  70 kg. The printed Equations 2 and 3 nevertheless write the denominator
  symbolically as `BW_median`, and the study medians are 62.0 kg (pregnant) and
  59.5 kg (non-pregnant). The model uses 70 kg, following the text and the
  Table 2 note.

* **Transit chain length.** Figure 1 draws a dose compartment transferring at
  `ktr` into a block of `n = 3` transit compartments, which then transfers at a
  separately estimated `ka` into the central compartment. The model therefore
  uses `ktr = 3 / MTT`, i.e. MTT spans the three `ktr`-governed transfers only,
  matching the convention of the co-author's
  `modellib("Ding_2024_amodiaquine")` (2 transits, `ktr = 2 / MTT`, separate
  `ka`). The alternative Savic-style reading `ktr = 4 / MTT` was tested against
  the published secondary parameters and is not distinguishable by them: because
  Cmax occurs after the last of six doses, the two readings differ by under 3%
  on median Cmax and under 2% on the day-7 concentration. No NONMEM control
  stream was available to settle the convention directly.

* **Box-Cox transformed random effect on F.** Results 3.1.2 describes the
  Box-Cox transformation as applying to the *inter-individual* variability of
  relative bioavailability, while Table 2 reports the single F variance term as
  *inter-occasion* variability (64.0% CV) and reports no separate F IIV row.
  The model encodes what Table 2 tabulates: one F variance term, carried across
  the two pooled dose occasions, with the Box-Cox shape `lambda = -0.690`
  applied to it in the Petersson 2009 form
  `eta_bc = (exp(lambda * eta) - 1) / lambda`. At `eta = 0` this returns 0 for
  any `lambda`, so the typical value of F is unaffected.

* **No inter-individual variability on CL/F, Q/F or Vp/F.** Equation 1 states
  that inter-individual variability was added exponentially to all PK
  parameters, but Table 2 of the final model reports variance terms only for
  MTT, Vc/F, ka and F. The model encodes the four terms that Table 2 reports;
  no variance is invented for the others. Almost all of the simulated exposure
  variability therefore arrives through the inter-occasion random effect on F.

* **Residual error scale.** Methods 2.10 says residual variability was
  evaluated "as additive, proportional, or combined additive and proportional
  errors on log-transformed concentrations", while Table 2 reports the retained
  combined model with a proportional term in per cent and an additive term in
  ng/mL, i.e. on the linear concentration scale. The model encodes
  `Cc ~ prop(propSd) + add(addSd)` on the linear scale, with the additive term
  converted from 12.0 ng/mL to 0.0120 ug/mL to match the model's `ug/mL`
  concentration unit.

* **Gestational age was significant but is not encoded.** Results 3.1.2 reports
  gestational age as a significant covariate on clearance (p < .01) and central
  volume (p < .05), but the model with binary pregnancy status had the lowest
  objective function value and was chosen as the final model; no
  gestational-age coefficient is printed. Gestational age is recorded in
  `covariatesDataExcluded` rather than in `covariateData`.

* **Dose50 value.** Methods 2.11 fixes Dose50 at 3.84 mg/kg, "taken from a
  large pooled analysis of lumefantrine" (reference 8, Kloprogge 2018). The
  sibling model `modellib("Kloprogge_2018_lumefantrine")` carries 3.86 mg/kg
  from its own Table 2. This model uses the 3.84 mg/kg value printed by
  Juma 2026.

* **Body-weight summaries conflict between Table 1 and Results 3.1.** Table 1
  reports a non-pregnant median body weight of 59.5 kg (45.0-89.0), while
  Results 3.1 quotes "non-pregnant women [median 62 (range: 40-87)]" in the
  per-trimester comparison. The virtual cohort uses the Table 1 value for the
  non-pregnant arm and the Results 3.1 values for the two pregnant arms, which
  are given only there.

* **Virtual cohort covariates.** Body weights are drawn log-normally around the
  published medians and truncated to the published ranges; the real joint
  distribution of weight, age and parasitaemia is not published. Age, baseline
  temperature and baseline parasitaemia were screened by the authors and not
  retained in the final model, so they are not simulated.

* **The 3rd-trimester versus 2nd-trimester ordering is not reproduced.** Table 3
  reports slightly higher Cmax and AUC in the third trimester than the second,
  whereas the model, whose only trimester-distinguishing input is body weight
  and whose allometry lowers exposure as weight rises, predicts the opposite
  ordering. The paper itself found no significant trimester effect, and the two
  published medians sit well inside each other's confidence intervals, so this
  is a limitation of using body weight as a trimester proxy rather than a
  transcription discrepancy.

* **Desbutyl-lumefantrine is not modelled.** The authors note this as a
  limitation of the study: the active metabolite was not measured and no
  metabolite model was built.
