---
title: "Fractal kinetics in population PK (Jung 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fractal kinetics in population PK (Jung 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 4.2
)
```

```{r setup}
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(PKNCA)
```

# Fractal kinetics applied to five population PK models

Jung 2023 (*Pharmaceutics* 15(1):304, doi:10.3390/pharmaceutics15010304) is a
methods paper with a concrete deliverable: it takes five previously fitted
population PK models, replaces one first-order rate constant in each with a
**fractal (Kopelman) rate**, re-estimates every model, and reports the full
converged parameter set for all five. This vignette implements the five
re-estimated (fractal) models and checks them.

The fractal modification replaces a constant first-order rate constant by a
time-dependent *instantaneous rate coefficient* (Jung 2023 Equation 1):

$$k(t) = \frac{q}{t^{\,h}}$$

where `q` carries the units of a first-order rate constant and `h` is a
unitless **heterogeneity exponent** bounded in `[0, 1]`. Setting `h = 0`
recovers ordinary constant-rate (Fick) kinetics; increasing `h` front-loads the
process, producing a burst followed by a long tail. The physical motivation is
a reaction on a fractal or otherwise non-homogeneous support, where the
effective rate falls as the readily accessible material is consumed.

In every one of the five cases the fractal rate is applied to an **absorption
or release** step, never to disposition.

```{r models}
mCase1 <- readModelDb("Jung_2023_donepezil_singledose")
mCase2 <- readModelDb("Jung_2023_donepezil_multipledose")
mCase3 <- readModelDb("Jung_2023_im_depot")
mCase4 <- readModelDb("Jung_2023_HL2351")
mCase5 <- readModelDb("Jung_2023_anakinra")
```

## The five cases

```{r case-table, echo = FALSE}
tibble::tibble(
  Case = 1:5,
  Model = c(
    "Jung_2023_donepezil_singledose",
    "Jung_2023_donepezil_multipledose",
    "Jung_2023_im_depot",
    "Jung_2023_HL2351",
    "Jung_2023_anakinra"
  ),
  Drug = c("donepezil", "donepezil (inferred)", "not disclosed", "HL2351", "anakinra"),
  Structure = c(
    "2-cmt, oral + patch, 2 transit",
    "2-cmt, oral + patch, 1 transit",
    "2-cmt IM, fast + slow (Savic transit)",
    "2-cmt QSS-TMDD + FcRn recycling",
    "1-cmt QSS-TMDD"
  ),
  `Fractal rate on` = c(
    "patch reservoir release",
    "patch reservoir release (reset per dose)",
    "fast IM release",
    "absorption site to plasma",
    "absorption from injection site"
  )
) |>
  knitr::kable(caption = "The five model cases of Jung 2023 (Section 2.3).")
```

Jung 2023 Table 2 reports the fit statistics. The fractal parameter improved
the objective function in every case, but by the corrected AIC it was worth the
extra parameter in only four of the five: Case 3's `h` estimated to 0.0268, an
exponent so close to zero that the model reduces to its own base model.

```{r table2, echo = FALSE}
tab2 <- tibble::tibble(
  Case          = 1:5,
  Subjects      = c(18L, 44L, 20L, 40L, 8L),
  Observations  = c(383L, 3024L, 339L, 472L, 93L),
  `OFV base`    = c(1443.70, 13977.10, 2155.43, 1556.43, 358.47),
  `OFV fractal` = c(1410.08, 13592.00, 2153.54, 1539.64, 350.13),
  `d OFV`       = c(-33.62, -385.10, -1.89, -16.79, -8.34),
  `d AICc`      = c(-31.47, -381.06, 2.55, -14.58, -3.12)
)
knitr::kable(tab2, caption = "Jung 2023 Table 2: performance of the fractal models.")
```

## What the fractal exponent does

This reproduces the mechanism illustrated in Jung 2023 Figures 2 and 3: the
instantaneous rate coefficient for a range of `h`, and the amount-remaining
profile it produces from a single depot. `h = 0` is a straight exponential
decay; larger `h` gives an increasingly sharp initial loss followed by a
markedly slower tail.

```{r fractal-shape, echo = FALSE}
q_rate <- 0.1
tgrid  <- seq(0.05, 24, by = 0.05)
shape <- expand.grid(h = seq(0, 1, by = 0.25), time = tgrid) |>
  mutate(
    rate = q_rate / time^h,
    # cumulative integral of q/t^h from 0 to t is q * t^(1-h) / (1-h);
    # at h = 1 the limit is q * log(t) with a lower cutoff.
    remaining = ifelse(
      h < 1,
      exp(-q_rate * time^(1 - h) / (1 - h)),
      exp(-q_rate * (log(time) - log(min(tgrid))))
    )
  )

ggplot(shape, aes(time, rate, colour = factor(h))) +
  geom_line(linewidth = 0.7) +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Instantaneous rate coefficient (1/h, log scale)",
    colour = "h",
    title = "Fractal rate coefficient k(t) = q / t^h, q = 0.1 /h"
  ) +
  theme_bw()
```

```{r fractal-remaining, echo = FALSE}
ggplot(shape, aes(time, remaining, colour = factor(h))) +
  geom_line(linewidth = 0.7) +
  labs(
    x = "Time (h)", y = "Fraction remaining in the depot", colour = "h",
    title = "Depot emptying under a fractal rate"
  ) +
  theme_bw()
```

# Source trace

Jung 2023's main text contains **no parameter values** -- Table 2 carries only
OFV, AIC and AICc. Every estimate in the five model files comes from the
Supplementary Materials, which contain Tables S1-S5 (one estimate table per
case) and Codes S1-S11 (the NONMEM control streams, base and fractal for each
case). Where the two disagree the control stream is authoritative, because its
`$THETA` block carries the converged values to six significant figures while
the tables are rounded to three; the two agree everywhere except the single
Case 3 residual-error discrepancy noted under Errata.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Case, ~Quantity, ~Source,
  1L, "Structural, IIV, residual error", "Table S1 (rounded), Code S3 $THETA / $OMEGA (converged)",
  1L, "ODE structure, fractal rate site", "Code S3 $MODEL / $DES / $PK; Figure S1",
  2L, "Structural, IIV, residual error", "Table S2 (rounded), Code S5 $THETA / $OMEGA (converged)",
  2L, "ODE structure, per-dose reset, tau", "Code S5 $MODEL / $DES / $PK; Figure S2",
  3L, "Structural, IIV, residual error", "Table S3 (rounded), Code S7 $THETA / $OMEGA (converged)",
  3L, "ODE structure, Savic transit chain", "Code S7 $MODEL / $DES / $PK; Figure S3",
  4L, "Structural, IIV, residual error", "Table S4 (rounded), Code S9 $THETA / $OMEGA (converged)",
  4L, "ODE structure, QSS-TMDD + FcRn", "Code S9 $MODEL / $DES / $PK / $ERROR; Figure S4",
  5L, "Structural, IIV, residual error", "Table S5 (rounded), Code S11 $THETA / $OMEGA (converged)",
  5L, "ODE structure, QSS-TMDD", "Code S11 $MODEL / $DES / $PK / $ERROR; Figure S5",
  NA_integer_, "Subject and observation counts", "Jung 2023 main text Table 2",
  NA_integer_, "Fit statistics (OFV / AIC / AICc)", "Jung 2023 main text Table 2",
  NA_integer_, "Case 1 study design and demographics", "Jung 2022, doi:10.3390/pharmaceutics14020244 (ref [3])",
  NA_integer_, "Case 4 / 5 study design and demographics", "Ngo 2020, doi:10.1002/psp4.12552 (ref [16])"
) |>
  knitr::kable(caption = "Where each model quantity comes from.")
```

Every `ini()` entry in the five model files additionally carries an in-file
comment naming its `$THETA` / `$OMEGA` index and the matching table row, so the
provenance of individual numbers is auditable at the line level.

# Verification

The checks below are deliberately built from values **printed in the paper**
(clearances from Tables S1-S5, the transit mean time, the QSS constants), never
from variables read back out of the model object. A gate assembled from the
model's own parameters cannot fail even when the model is wrong.

Jung 2023 reports no non-compartmental summary statistics -- no Cmax, Tmax, AUC
or half-life tables -- so there is nothing to compare a simulated NCA against.
The NCA below therefore serves a different purpose: for the three linear cases
it tests the dose-exposure identity `AUC(0-inf) = Dose / CL` against the
*printed* clearance, which is a genuine end-to-end check of the ODE wiring, the
bioavailability routing and the concentration unit scaling all at once.

## Cases 1 and 2: donepezil, oral and transdermal

Typical-value profiles (inter-individual variability suppressed with
`rxode2::zeroRe()`) for the single-dose model.

```{r case1-profile}
ev1_oral <- rxode2::et(amt = 10, cmt = "depot_oral", time = 0) |>
  rxode2::et(seq(0, 312, by = 1), cmt = "central")
ev1_patch <- rxode2::et(amt = 108, cmt = "depot_td", time = 0) |>
  rxode2::et(seq(0, 312, by = 1), cmt = "central")

s1_oral  <- rxode2::rxSolve(rxode2::zeroRe(mCase1), ev1_oral,  returnType = "data.frame")
s1_patch <- rxode2::rxSolve(rxode2::zeroRe(mCase1), ev1_patch, returnType = "data.frame")

case1 <- bind_rows(
  mutate(s1_oral,  arm = "Oral 10 mg"),
  mutate(s1_patch, arm = "Patch 108 mg")
)

ggplot(case1, aes(time, Cc, colour = arm)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time (h)", y = "Donepezil (ng/mL)", colour = NULL,
       title = "Case 1: single-dose donepezil, typical-value profiles") +
  theme_bw()
```

The oral arm is a plain first-order absorption into a linear two-compartment
disposition, so its total exposure must satisfy `AUC = Dose / CL` using the
clearance printed in Table S1 (9.68 L/h). Dose is 10 mg and concentrations are
ng/mL, so the expected AUC is `10 mg / 9.68 L/h = 1.0331 mg*h/L = 1033.1
ng*h/mL`.

```{r case1-auc-gate}
# Trapezoidal AUC to 312 h plus the terminal tail, from the typical-value solve.
auc_with_tail <- function(d) {
  a_obs <- sum(diff(d$time) * (head(d$Cc, -1) + tail(d$Cc, -1)) / 2)
  # terminal slope from the last quarter of the profile
  tailrows <- tail(d, max(10L, floor(nrow(d) / 4)))
  lz <- -stats::coef(stats::lm(log(Cc) ~ time, data = tailrows))[["time"]]
  a_obs + tail(d$Cc, 1) / lz
}

cl_printed_case1 <- 9.68           # Jung 2023 Table S1, fractal column, CL (L/h)
auc_expected1    <- 10 / cl_printed_case1 * 1000   # mg / (L/h) -> ng*h/mL
auc_observed1    <- auc_with_tail(s1_oral)

c(expected = auc_expected1, observed = auc_observed1,
  pct_diff = 100 * (auc_observed1 - auc_expected1) / auc_expected1)

stopifnot(abs(auc_observed1 - auc_expected1) / auc_expected1 < 0.02)
```

Case 2 is the multiple-dose model, and the only one of the five whose fractal
clock **restarts at every dose**. The control stream (Code S5) tracks the time
of the most recent dosing record and forms the release coefficient from the
elapsed time since it, so each new patch begins its release profile afresh
rather than continuing to decay on an absolute clock. The check below is the
structural one: the fraction of the reservoir released in the first 24 h must
be the same for every patch.

```{r case2-reset}
ev2 <- rxode2::et(amt = 10, cmt = "depot_oral", time = 0, ii = 24, addl = 6) |>
  rxode2::et(amt = 108, cmt = "depot_td", time = 168) |>
  rxode2::et(amt = 108, cmt = "depot_td", time = 888) |>
  rxode2::et(amt = 108, cmt = "depot_td", time = 1608) |>
  rxode2::et(seq(0, 2496, by = 2), cmt = "central")

s2 <- rxode2::rxSolve(rxode2::zeroRe(mCase2), ev2, returnType = "data.frame")

released_24h <- vapply(c(168, 888, 1608), function(td) {
  w <- s2[s2$time >= td & s2$time <= td + 24, ]
  1 - tail(w$depot_td, 1) / w$depot_td[1]
}, numeric(1))

released_24h

# Every patch must release the same fraction in its first 24 h. If the clock ran
# on absolute time instead of time-after-dose, the later patches would release
# dramatically less and this spread would be large.
stopifnot(diff(range(released_24h)) < 1e-6)
```

```{r case2-profile, echo = FALSE}
ggplot(s2, aes(time, Cc)) +
  geom_line(linewidth = 0.6) +
  geom_vline(xintercept = c(168, 888, 1608), linetype = 2, colour = "grey50") +
  labs(x = "Time (h)", y = "Donepezil (ng/mL)",
       title = "Case 2: oral titration then three patches (dashed = patch application)") +
  theme_bw()
```

## Case 3: controlled-release intramuscular depot

Case 3 splits the dose between a fast-release depot given the fractal rate and a
slow-release arm routed through a Savic transit chain. The transit chain is the
part with a hard, printed target: the chain's **input rate** into the slow depot
is a gamma density in elapsed time whose mode is the mean transit time, so it
must peak at `t = MTT`. Table S3 prints `MTT = 136 h`.

The quantity to test is the input rate `inpt`, not the `depot2` state. `depot2`
accumulates that input while simultaneously draining at `ka_slow`, so its own
maximum falls well after `MTT` (about 173 h here) and would not test the transit
chain at all.

```{r case3-transit}
ev3 <- rxode2::et(amt = 1000, cmt = "depot", time = 0) |>
  rxode2::et(seq(0, 672, by = 1), cmt = "central")
s3 <- rxode2::rxSolve(rxode2::zeroRe(mCase3), ev3, returnType = "data.frame")

mtt_printed  <- 136           # Jung 2023 Table S3, fractal column, MTT (h)
t_peak_input <- s3$time[which.max(s3$inpt)]

c(mtt_printed = mtt_printed, simulated_input_peak_h = t_peak_input)

# The transit input rate must peak at t = MTT. The observation grid is 1 h, so
# allow the grid resolution plus one step.
stopifnot(abs(t_peak_input - mtt_printed) <= 2)
```

The transit chain also has a conservation property: integrated over all time
the input must deliver exactly the slow-arm fraction of the dose,
`(1 - FRAC) * Dose`, with `FRAC = 0.172` printed in Table S3.

```{r case3-transit-mass}
frac_printed <- 0.172         # Jung 2023 Table S3, fractal column, FRAC
dose3 <- 1000
delivered <- sum(diff(s3$time) * (head(s3$inpt, -1) + tail(s3$inpt, -1)) / 2)
expected3 <- (1 - frac_printed) * dose3

c(expected = expected3, delivered = delivered,
  pct_diff = 100 * (delivered - expected3) / expected3)

stopifnot(abs(delivered - expected3) / expected3 < 0.01)
```

```{r case3-profile, echo = FALSE}
ggplot(s3, aes(time, Cc)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time (h)", y = "Concentration (ug/L)",
       title = "Case 3: controlled-release IM depot, typical-value profile") +
  theme_bw()
```

## Cases 4 and 5: target-mediated disposition

Both TMDD cases carry the drug as a **total** amount and recover the free
concentration through a quasi-steady-state quadratic. Two properties must hold
identically, whatever the parameter values: free concentration never exceeds
total, and as the total rises far above the target capacity the free fraction
approaches one. These are structural checks on the QSS algebra rather than on
any single estimate.

```{r case45-qss}
NMOL_PER_MG_HL2351 <- 1e6 / 97000   # MW ~97 kDa -> 1 mg = 10.31 nmol
wt <- 70

case4 <- lapply(c(1, 4, 12), function(mgkg) {
  ev <- rxode2::et(amt = mgkg * wt * NMOL_PER_MG_HL2351, cmt = "depot", time = 0) |>
    rxode2::et(seq(0, 672, by = 2), cmt = "central")
  rxode2::rxSolve(rxode2::zeroRe(mCase4), ev, returnType = "data.frame") |>
    mutate(dose = paste0(mgkg, " mg/kg"))
}) |> bind_rows()

# Cc is the FREE concentration; central/vc is the TOTAL. Free <= total always.
case4chk <- case4 |>
  mutate(total = central / 11.2913) |>     # Vc printed in Table S4 (L)
  filter(total > 0)

stopifnot(all(case4chk$Cc <= case4chk$total * (1 + 1e-8)))
stopifnot(all(is.finite(case4$Cc)))

# At high total concentration the free fraction must approach 1.
high <- case4chk |> filter(total > 50)
round(range(high$Cc / high$total), 3)
stopifnot(min(high$Cc / high$total) > 0.75)
```

```{r case4-profile, echo = FALSE}
ggplot(case4, aes(time, Cc, colour = dose)) +
  geom_line(linewidth = 0.7) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Free HL2351 (nmol/L, log scale)", colour = NULL,
       title = "Case 4: HL2351 single ascending SC doses, typical-value profiles") +
  theme_bw()
```

```{r case5-profile}
NMOL_PER_MG_ANAKINRA <- 1e6 / 17300   # MW ~17.3 kDa
ev5 <- rxode2::et(amt = 100 * NMOL_PER_MG_ANAKINRA, cmt = "depot", time = 0) |>
  rxode2::et(seq(0, 48, by = 0.25), cmt = "central")
s5 <- rxode2::rxSolve(rxode2::zeroRe(mCase5), ev5, returnType = "data.frame")

# Anakinra 100 mg SC is reported in the literature to peak around 1 ug/mL.
# 1 ug/mL / 17.3 kDa = 57.8 nmol/L. This is an order-of-magnitude sanity gate on
# the unit handling, not a published NCA comparison.
cmax5_ugml <- max(s5$Cc) * 17300 / 1e6
c(cmax_nmol_L = max(s5$Cc), cmax_ug_mL = cmax5_ugml,
  tmax_h = s5$time[which.max(s5$Cc)])

stopifnot(cmax5_ugml > 0.3, cmax5_ugml < 3)
```

```{r case5-plot, echo = FALSE}
ggplot(s5, aes(time, Cc)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time (h)", y = "Free anakinra (nmol/L)",
       title = "Case 5: anakinra 100 mg SC, typical-value profile") +
  theme_bw()
```

## Setting h to zero recovers constant-rate kinetics

The strongest available check on the fractal implementation itself is Jung
2023's own defining property: at `h = 0` the instantaneous coefficient
`q / t^h` collapses to the constant `q`, so the fractal model must reproduce
its base-model absorption exactly. Case 5 has the simplest absorption path, so
the collapse is checked there against an independently written one-compartment
first-order-absorption solution.

```{r h-zero-collapse}
m5_h0 <- mCase5 |> rxode2::zeroRe()
s5_h0 <- rxode2::rxSolve(m5_h0, ev5, params = c(h_abs = 0), returnType = "data.frame")

# Independent reference: with h = 0 the depot empties at the constant rate ka,
# printed in Table S5 as 0.47 /h (Code S11 $THETA 3 = 0.469732).
ka0 <- 0.469732
dose5 <- 100 * NMOL_PER_MG_ANAKINRA
depot_ref <- dose5 * exp(-ka0 * s5_h0$time)

max_rel_err <- max(abs(s5_h0$depot - depot_ref)) / dose5
max_rel_err
stopifnot(max_rel_err < 1e-6)
```

The depot trajectory matches a hand-written exponential to better than one part
in a million, which confirms both that `h` enters the rate as Equation 1
specifies and that the printed `Ka` is the coefficient `q`.

## Non-compartmental analysis

```{r pknca}
nca_input <- bind_rows(
  s1_oral |> transmute(id = 1L, treatment = "Case 1 oral 10 mg", time, Cc),
  s3      |> transmute(id = 2L, treatment = "Case 3 IM 1000 ug", time, Cc)
) |>
  filter(!is.na(Cc))

dose_df <- tibble::tibble(
  id        = c(1L, 2L),
  treatment = c("Case 1 oral 10 mg", "Case 3 IM 1000 ug"),
  time      = 0,
  amt       = c(10, 1000)
)

conc_obj <- PKNCA::PKNCAconc(nca_input, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

nca_res <- suppressWarnings(
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
)

knitr::kable(
  summary(nca_res),
  caption = "Simulated NCA of the two linear typical-value profiles."
)
```

The dose-exposure identity, checked against the clearances printed in Tables S1
and S3. Both cases are linear in dose, so `AUC(0-inf)` must equal `Dose / CL`.

```{r nca-vs-printed}
nca_tbl <- as.data.frame(nca_res)

auc_sim <- nca_tbl |>
  filter(PPTESTCD == "aucinf.obs") |>
  select(treatment, simulated = PPORRES)

expected <- tibble::tibble(
  treatment = c("Case 1 oral 10 mg", "Case 3 IM 1000 ug"),
  cl_printed = c(9.68, 0.175),                 # Tables S1 and S3, fractal column (L/h)
  dose       = c(10, 1000),                    # mg and ug
  expected   = c(10 / 9.68 * 1000, 1000 / 0.175)
)

comparison <- left_join(expected, auc_sim, by = "treatment") |>
  mutate(`% difference` = 100 * (simulated - expected) / expected)

comparison |>
  rename(
    "Profile"            = treatment,
    "Printed CL (L/h)"   = cl_printed,
    "Dose"               = dose,
    "Expected AUC"       = expected,
    "Simulated AUC"      = simulated
  ) |>
  knitr::kable(
    digits = c(0, 4, 0, 1, 1, 2),
    caption = "AUC(0-inf) against Dose / CL using the clearances printed by Jung 2023."
  )

stopifnot(all(abs(comparison$`% difference`) < 5))
```

Both profiles reproduce `Dose / CL` to within a few percent, the residue being
the numerical tail extrapolation rather than a structural discrepancy. Because
the expected column is built from the *printed* clearance and the dose, and the
simulated column from a numerical solve of the ODE system, this comparison
would fail on a mis-transcribed clearance, a wrong bioavailability route or a
concentration unit error.

# Assumptions and deviations

- **The drug in Case 2 is inferred, not stated.** Jung 2023 never names it.
  Donepezil is adopted on the evidence set out in the model file's
  `population$notes` (same group and oral-plus-patch programme as Case 1;
  identical `A(2)/(VC/1000)` ng/mL concentration scaling in Codes S3 and S5;
  disposition estimates close to Case 1's). Against it: the paper does not say
  so, and three patches over 2496 h is about 35 days per patch, which does not
  match the weekly patch of ref [3]. The parameter values and the structure do
  not depend on the identification. Adopted per operator sidecar
  `oare_PMC9867137` request-001 q2.
- **The drug in Case 3 is not disclosed at all**, so the model file is named
  `Jung_2023_im_depot` rather than `author_year_drug`. The data are described
  only as in-house.
- **Dose amounts for Cases 2 and 3 are not reported.** Jung 2023 states only
  that two different amounts were dosed for the oral and patch routes in Case 2,
  and four different amounts in Case 3. The amounts used in this vignette (the
  Case 1 oral 10 mg and 108 mg patch strengths; 1000 ug for Case 3) are
  illustrative choices for the simulation, not values from the paper. All three
  linear models scale proportionally with dose, so the gates above are
  unaffected.
- **Species is stated nowhere in Jung 2023.** Cases 1, 4 and 5 are pinned to
  human clinical studies through refs [3] and [16]. Cases 2 and 3 are recorded
  as human by inference from their four sibling cases.
- **Case 3's residual error conflicts between table and code, resolved in favour
  of the code.** Table S3's fractal column prints both an additive error (0.133)
  and a proportional error (0.323), but Code S7's `$ERROR` block is
  `W = SQRT(THETA(11)**2 * IPRED**2)` -- proportional only, with a single theta
  (0.112738, labelled "Add.err"). A one-theta error block cannot produce two
  reported values, the base stream S6 has the identical one-theta proportional
  form, and all ten other S7 thetas reproduce Table S3 exactly. The model file
  implements the control stream.
- **Case 2's per-dose reset is implemented with `tad()` on the patch reservoir.**
  Code S5 resets its elapsed-time clock on any dosing record, which is a broader
  trigger than resetting on patch doses alone. The two are equivalent for this
  study's sequential design (oral titration completes before the first patch),
  and the reset is verified numerically above. The release flux is computed
  inside a guard because rxode2's `tad()` returns `NA` before the first dose to a
  compartment and `NA * 0` is `NA`, which would silently poison the whole solve.
- **No onset-time regulariser is added where the authors did not specify one.**
  Only Case 2 carries a `TAU` (1e-7 h, hardcoded in Code S5). Because `h < 1`
  makes the `t^-h` singularity integrable, the unregularised cases solve without
  one; adding a regulariser would introduce an unsourced parameter.
- **No published NCA to compare against.** Jung 2023 reports OFV, AIC and AICc
  and shows GOF, VPC and NPDE figures, but tabulates no Cmax, Tmax, AUC or
  half-life. The NCA section therefore validates the dose-exposure identity
  against printed clearances instead of reproducing a published table.
- **Only the fractal models are shipped.** Each case has a base and a fractal
  variant (Codes S2/S3, S4/S5, S6/S7, S8/S9, S10/S11). The fractal model is the
  paper's contribution and the final model of each pair; the base models appear
  here only through their fit statistics in the Table 2 reproduction above.
- **Case 3's fractal exponent is effectively zero.** `h = 0.0268` with 124% CV
  and a positive delta-AICc: by the authors' own criterion the fractal rate is
  not warranted for this case. It is implemented as published, but users should
  not read it as evidence of fractal behaviour.
