---
title: "Donepezil oral and transdermal patch (Jung 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Donepezil oral and transdermal patch (Jung 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Jung_2022_donepezil"))$reference`
* Description: `r rxode2::rxode(readModelDb("Jung_2022_donepezil"))$description`
* Article: <https://doi.org/10.3390/pharmaceutics14020244>
* Control stream with unrounded estimates (Code S2 of a later paper by the same
  group): <https://doi.org/10.3390/pharmaceutics15010304>

Jung 2022 built one integrated population PK model for a once-daily 10 mg oral
donepezil tablet and a once-weekly donepezil transdermal patch, then used it to
find the patch strengths whose simulated steady-state AUC and Cmax are
bioequivalent to the tablet. The two routes share one central compartment, one
peripheral compartment and one clearance. Oral drug is absorbed from a gut
depot at first-order `Ka`; patch drug leaves a skin reservoir at first-order
`Kt` and passes through two transit compartments with the same `Kt` before
reaching plasma (Jung 2022 Eqs. 1, 2 and 4, Figure 1).

The amount that enters the skin reservoir is not the nominal patch strength.
Jung 2022 Eq. 3 fits the in-vitro dissolution profile as

$$\text{Dissolution}(D) = \frac{78.257\,D}{D + 8.481}\ \%\ \text{of patch dose},$$

with `D` the wear time in hours, and scales it by 0.74, the ratio of the drug
released in the clinical study (measured from the residue left in used patches)
to the in-vitro release. At the one-week wear used in the study and in every
simulation of the paper this gives
`0.74 * 0.78257 * 168 / (168 + 8.481) = 0.5513`, carried in the model as the
fixed bioavailability `f(depot_td)`. Dose the **nominal** patch strength.

The same authors refit this model unchanged as the "base model" of Case 1 in a
2023 methods paper on fractal absorption kinetics, and published its NONMEM
control stream (Code S2) there. Code S2 reproduces Jung 2022 Table 3 exactly
(identical OFV 1443.703, every estimate rounding to the Table 3 value), so the
unrounded Code S2 values are used here. The fractal refit of the same data
ships separately as `Jung_2023_donepezil_singledose`.

# Population

Twelve healthy male volunteers aged 18-45 years were enrolled in a randomised,
open-label, two-treatment, two-sequence, two-period crossover study
(TL/WZ/19/001141). Nine completed both periods and form the analysis set
(Jung 2022 Section 3.1, Table 1): age 24-33 years (median 30.0), weight
55.7-80.9 kg (median 63.1), BMI 19.8-27.3 kg/m^2, all male and all Asian. The
test period was one 108 mg / 96 cm^2 patch worn on the torso or back for one
week; the reference period was 10 mg Aricept once daily for seven days, with a
washout of at least 21 days. Blood was sampled to 312 h. The model has no
covariates.

```{r population}
readModelDb("Jung_2022_donepezil")()$population |> str()
```

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `d/dt(depot_oral) = -ka * depot_oral` | n/a | Jung 2022 Eq. 1; Code S2 `$DES` DADT(1) |
| `d/dt(depot_td)`, `d/dt(transit1)`, `d/dt(transit2)` (one `Kt`) | n/a | Jung 2022 Eq. 2; Code S2 `$DES` DADT(4)-DADT(6) |
| `d/dt(central)`, `d/dt(peripheral1)`, `kel = CL/Vc`, `kcp = Q/Vc`, `kpc = Q/Vp` | n/a | Jung 2022 Eq. 4; Code S2 `$PK` / `$DES` |
| `f(depot_td) = 0.74 * 0.78257 * 168 / (168 + 8.481)` | 0.5513 | Jung 2022 Eq. 3 with the one-week wear of Section 2.1 |
| `Cc = 1000 * central / vc` (ng/mL) | n/a | Code S2 `$ERROR`, `IPRED = A(2)/(VC/1000)` |
| `lka` | log(0.0496711) 1/h | Table 3 Ka 0.0497; Code S2 THETA(1) |
| `lcl` | log(10.0268) L/h | Table 3 CL 10; Code S2 THETA(2) |
| `lvc` | log(26.2189) L | Table 3 Vc 26.2; Code S2 THETA(3) |
| `lvp` | log(562.037) L | Table 3 Vp 562; Code S2 THETA(4) |
| `lq` | log(15.6292) L/h | Table 3 Q 15.6; Code S2 THETA(5) |
| `lktr` | log(0.0270191) 1/h | Table 3 Kt 0.027; Code S2 THETA(6) |
| `etalka` | 0.00968106 (9.9% CV) | Table 3; Code S2 OMEGA(1) |
| `etalcl` | 0.130076 (37.3% CV) | Table 3; Code S2 OMEGA(2) |
| `etalvc` | 0.197923 (46.8% CV) | Table 3; Code S2 OMEGA(3) |
| `etalktr` | 0.020045 (14.2% CV) | Table 3; Code S2 OMEGA(4) |
| `addSd` | 2.89074 ng/mL | Table 3 additive error 2.89; Code S2 THETA(7) |
| `propSd` | 0.0795173 | Table 3 proportional error 0.0795; Code S2 THETA(8) |

The Table 3 "IIV" column is a variance: `sqrt(exp(omega) - 1)` recovers its
printed CV% for every row (for example `sqrt(exp(0.13) - 1) = 37.3%`).

```{r omega-scale}
mod <- readModelDb("Jung_2022_donepezil")
ui <- rxode2::rxode2(mod)
omega_cv <- 100 * sqrt(exp(diag(ui$omega)) - 1)
published_cv <- c(etalka = 9.9, etalcl = 37.3, etalvc = 46.8, etalktr = 14.2)
round(rbind(model = omega_cv[names(published_cv)], published = published_cv), 1)
stopifnot(max(abs(omega_cv[names(published_cv)] - published_cv)) < 0.1)
```

# Structural checks on the typical subject

Both routes are linear, so the total exposure of the typical subject is fixed by
clearance alone: `AUC(0,inf) = F * Dose / CL`. The oral arm (seven 10 mg doses,
`F = 1`) and the patch arm (108 mg, `F = 0.5513`) are integrated with tight
tolerances on a grid long enough for the 562 L peripheral volume to empty.

```{r typical-identity}
typ <- rxode2::zeroRe(mod)
f_td <- exp(ui$theta[["lfdepot_td"]])
cl_typ <- exp(ui$theta[["lcl"]])
grid <- sort(unique(c(seq(0, 200, by = 0.1), seq(200, 4000, by = 1))))

ev_oral_typ <- rxode2::et(amt = 10, cmt = "depot_oral", time = 0, ii = 24, addl = 6) |>
  rxode2::et(grid, cmt = "central")
ev_patch_typ <- rxode2::et(amt = 108, cmt = "depot_td", time = 0) |>
  rxode2::et(grid, cmt = "central")

trap <- function(t, y) sum(diff(t) * (head(y, -1) + tail(y, -1)) / 2)
solve_typ <- function(ev) {
  as.data.frame(rxode2::rxSolve(typ, ev, rtol = 1e-10, atol = 1e-12, maxsteps = 1e6))
}
s_oral_typ <- solve_typ(ev_oral_typ)
s_patch_typ <- solve_typ(ev_patch_typ)

identity <- tibble::tibble(
  arm = c("Oral 10 mg x 7", "Patch 108 mg"),
  expected = c(70 / cl_typ, f_td * 108 / cl_typ) * 1000,
  simulated = c(trap(s_oral_typ$time, s_oral_typ$Cc), trap(s_patch_typ$time, s_patch_typ$Cc))
) |>
  mutate(pct_diff = 100 * (simulated - expected) / expected)
knitr::kable(identity, digits = c(0, 1, 1, 4),
             caption = "Typical-subject AUC(0,inf) (ng*h/mL) against F * Dose / CL.")
# Deterministic solve against its own closed form; the residual (measured
# ~0.001%) is the trapezoid on the grid plus the concentration left at 4000 h.
# Dropping the Eq. 3 fraction or a digit of CL moves it by tens of percent.
stopifnot(max(abs(identity$pct_diff)) < 0.02)
```

The patch identity is also an independent check of Eq. 3. The typical patch
AUC is `r round(identity$simulated[2])` ng*h/mL; the observed mean AUCinf of the
nine patch subjects is 5909 ng*h/mL (Jung 2022 Table 2). Dosing the nominal
108 mg without the Eq. 3 fraction would predict about 10 800 ng*h/mL, so the
delivered fraction is a load-bearing part of the model, not a detail.

```{r eq3-check}
auc_patch_obs <- 5909.34 # Jung 2022 Table 2, Patch (0-312 h), mean AUCinf
stopifnot(
  abs(identity$simulated[2] / auc_patch_obs - 1) < 0.05,
  # Nominal dosing (F = 1) would be ~1.8-fold too high.
  108 / cl_typ * 1000 / auc_patch_obs > 1.7
)
```

# Virtual cohort and simulation

The study cohort is reproduced as two arms of 200 virtual subjects each (the
model has no covariates, so the arms differ only in their random effects): the
reference regimen of 10 mg once daily for seven days and the test regimen of a
single 108 mg patch, both observed to 312 h as in the study.

```{r cohort}
rxode2::rxSetSeed(20220120)
n_arm <- 200
obs_times <- sort(unique(c(seq(0, 24, by = 0.5), seq(25, 312, by = 1))))

make_arm <- function(n, treatment, id_offset, dose_rows) {
  ids <- id_offset + seq_len(n)
  doses <- tidyr::crossing(id = ids, dose_rows) |>
    mutate(evid = 1L)
  obs <- tidyr::crossing(id = ids, time = obs_times) |>
    mutate(amt = 0, cmt = "central", evid = 0L)
  bind_rows(doses, obs) |>
    mutate(treatment = treatment) |>
    arrange(id, time, desc(evid))
}

events <- bind_rows(
  make_arm(n_arm, "Oral", 0L,
           tibble::tibble(time = 24 * (0:6), amt = 10, cmt = "depot_oral")),
  make_arm(n_arm, "Patch", n_arm,
           tibble::tibble(time = 0, amt = 108, cmt = "depot_td"))
)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))

sim <- rxode2::rxSolve(mod, events = events, keep = "treatment", maxsteps = 1e6) |>
  as.data.frame()
stopifnot(!anyNA(sim$Cc))
```

# Replicate published figures

```{r figure-2, fig.width = 7, fig.height = 6}
# Replicates Figure 2 of Jung 2022: VPC of donepezil after oral (upper) and
# transdermal patch (lower) administration.
sim |>
  group_by(treatment, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~treatment, ncol = 1, scales = "free_y") +
  labs(x = "Time (h)", y = "Donepezil (ng/mL)",
       title = "Simulated 5th, 50th and 95th percentiles",
       caption = "Replicates Figure 2 of Jung 2022 (200 virtual subjects per arm).") +
  theme_bw()
```

# PKNCA validation

Jung 2022 Table 2 reports arithmetic means for three analyses: the first oral
dose (0-24 h), the full oral course (0-312 h) and the patch (0-312 h). The
oral profile is entered twice, once truncated at 24 h with only the first
dose, so that each analysis is its own PKNCA group.

```{r pknca}
conc_all <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)

conc_nca <- bind_rows(
  conc_all |> filter(treatment == "Oral", time <= 24) |> mutate(analysis = "Oral (0-24 h)"),
  conc_all |> filter(treatment == "Oral") |> mutate(analysis = "Oral (0-312 h)"),
  conc_all |> filter(treatment == "Patch") |> mutate(analysis = "Patch (0-312 h)")
)
# Time-zero anchor for every subject (pre-dose Cc = 0 for extravascular dosing).
conc_nca <- bind_rows(
  conc_nca,
  conc_nca |> distinct(id, analysis) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, analysis, time, .keep_all = TRUE) |>
  arrange(analysis, id, time)

dose_all <- events |>
  filter(evid == 1) |>
  select(id, time, amt, treatment)
dose_nca <- bind_rows(
  dose_all |> filter(treatment == "Oral", time == 0) |> mutate(analysis = "Oral (0-24 h)"),
  dose_all |> filter(treatment == "Oral") |> mutate(analysis = "Oral (0-312 h)"),
  dose_all |> filter(treatment == "Patch") |> mutate(analysis = "Patch (0-312 h)")
)

conc_obj <- PKNCA::PKNCAconc(conc_nca, Cc ~ time | analysis + id)
dose_obj <- PKNCA::PKNCAdose(dose_nca, amt ~ time | analysis + id)

intervals <- bind_rows(
  data.frame(analysis = "Oral (0-24 h)", start = 0, end = 24,
             cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = FALSE, half.life = FALSE),
  data.frame(analysis = c("Oral (0-312 h)", "Patch (0-312 h)"), start = 0, end = Inf,
             cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE)
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

## Comparison against published NCA

Jung 2022 Table 2 reports means, so the simulated results are summarised as
arithmetic means before comparison. `AUCinf` of the first oral dose (575 ng*h/mL
in the paper) is extrapolated from 24 h of data with a 22 h half-life estimated
within that window and is not compared.

```{r nca-compare}
sim_mean <- as.data.frame(nca_res$result) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "aucinf.obs", "half.life")) |>
  group_by(analysis, PPTESTCD) |>
  summarise(value = mean(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = value)

published <- tibble::tribble(
  ~analysis,         ~cmax, ~tmax,  ~auclast, ~aucinf.obs, ~half.life,
  "Oral (0-24 h)",   20.26,   3.11,   286.62,          NA,        NA,
  "Oral (0-312 h)",  53.86, 146.33,  6111.08,     6873.40,     37.74,
  "Patch (0-312 h)", 28.62, 106.67,  5285.59,     5909.34,     71.17
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_mean,
  reference = published,
  by = "analysis",
  units = c(cmax = "ng/mL", tmax = "h", auclast = "ng*h/mL",
            aucinf.obs = "ng*h/mL", half.life = "h"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Simulated (mean of 200 per arm) vs. published (mean of 9) NCA. * differs from reference by >20%.")
```

```{r nca-gates}
num <- function(analysis, code) {
  v <- sim_mean[[code]][sim_mean$analysis == analysis]
  if (length(v) != 1L) stop("no unique row for ", analysis, " / ", code)
  v
}
pub <- function(analysis, code) {
  v <- published[[code]][published$analysis == analysis]
  if (length(v) != 1L) stop("no unique row for ", analysis, " / ", code)
  v
}
pct <- function(analysis, code) 100 * (num(analysis, code) / pub(analysis, code) - 1)

# Exposure gates on the centre of a 200-subject cohort. The CV of individual
# AUC is ~37% (the IIV on CL), so the standard error of a 200-subject mean is
# ~3%; the 15% bound sits > 4 SE out and still catches a mis-transcribed CL,
# a missing Eq. 3 fraction (+80%) or a wrong unit.
stopifnot(
  abs(pct("Oral (0-312 h)", "aucinf.obs")) < 15,
  abs(pct("Patch (0-312 h)", "aucinf.obs")) < 15,
  abs(pct("Patch (0-312 h)", "cmax")) < 20,
  # Patch Tmax is set by the three-step Kt chain (mean transit 3 / Kt = 111 h).
  abs(pct("Patch (0-312 h)", "tmax")) < 20
)
```

The patch arm agrees with Table 2 on every row, and the AUCs of both routes
agree to within 10%. The simulated mean AUCs sit about 6-7% above the typical
values of the identity check above, which is the expected gap between the mean
and the median of a log-normal exposure: `exp(0.130 / 2) = 1.067` for the IIV
on CL. Two rows differ more:

* **Oral Cmax** is 15-20% below the observed means, both after the first dose
  and at the end of the course, while the oral Tmax and AUC agree. The
  estimated `Ka` (0.0497 1/h) is slow for an immediate-release tablet; the
  authors note estimation difficulties from flip-flop kinetics, and a slow
  absorption constant flattens the oral peak more than it shifts its timing.
  The model is shipped as estimated.
* **Oral half-life** (starred). The oral course ends at 144 h, so the 0-312 h
  window holds only 168 h of washout, over which the 562 L peripheral
  compartment is still returning drug. The published 37.7 h (and 22.3 h from the
  first 24 h alone) are distribution-phase slopes from sparse sampling; the
  simulated value comes from a denser grid whose last points lie further into
  the terminal phase. The patch half-life, fitted after a longer absorption
  tail, agrees.

# Equivalent patch dose (Table 4 and Figure 3)

Jung 2022 simulated 10 mg oral donepezil once daily against once-weekly
patches and compared AUC and Cmax over the steady-state week 672-840 h. The
typical-subject ratios are deterministic, and reproduce the published point
estimates (which come from a 100-per-arm stochastic simulation):

```{r table-4}
ss_grid <- seq(672, 840, by = 0.25)
ss_solve <- function(ev) {
  s <- as.data.frame(rxode2::rxSolve(typ, ev, rtol = 1e-10, atol = 1e-12, maxsteps = 1e6))
  s[s$time >= 672 & s$time <= 840, ]
}
ref_ss <- ss_solve(rxode2::et(amt = 10, cmt = "depot_oral", time = 0, ii = 24, addl = 34) |>
                     rxode2::et(ss_grid, cmt = "central"))
patch_ratio <- function(dose) {
  s <- ss_solve(rxode2::et(amt = dose, cmt = "depot_td", time = 0, ii = 168, addl = 4) |>
                  rxode2::et(ss_grid, cmt = "central"))
  c(auc = 100 * trap(s$time, s$Cc) / trap(ref_ss$time, ref_ss$Cc),
    cmax = 100 * max(s$Cc) / max(ref_ss$Cc))
}
r114 <- patch_ratio(114)
r146 <- patch_ratio(146)

tab4 <- tibble::tibble(
  patch = c("114 mg / 101.3 cm2", "114 mg / 101.3 cm2", "146 mg / 129.8 cm2", "146 mg / 129.8 cm2"),
  metric = c("AUC", "Cmax", "AUC", "Cmax"),
  published_ratio = c(91.16, 86.66, 115.63, 108.77),
  published_ci90 = c("85.61-97.06", "82.07-91.51", "108.62-123.09", "102.98-114.88"),
  typical_ratio = c(r114[["auc"]], r114[["cmax"]], r146[["auc"]], r146[["cmax"]])
) |>
  mutate(pct_diff = 100 * (typical_ratio / published_ratio - 1))

tab4 |>
  dplyr::rename(
    "Patch" = patch, "Metric" = metric, "Published ratio (%)" = published_ratio,
    "Published 90% CI" = published_ci90, "Typical-subject ratio (%)" = typical_ratio,
    "% diff" = pct_diff
  ) |>
  knitr::kable(digits = 2, caption = "Replicates Table 4 of Jung 2022 (test patch / reference 10 mg oral daily).")

# Deterministic quantities. Every typical-subject ratio lies inside the published
# 90% CI; the AUC ratios are within 2% of the published point estimates. A
# mis-transcribed Kt, Eq. 3 constant or unit moves them by far more.
ci <- do.call(rbind, lapply(strsplit(tab4$published_ci90, "-"), as.numeric))
stopifnot(
  all(tab4$typical_ratio > ci[, 1] & tab4$typical_ratio < ci[, 2]),
  max(abs(tab4$pct_diff[tab4$metric == "AUC"])) < 2.5,
  max(abs(tab4$pct_diff)) < 6
)
```

```{r figure-3, fig.width = 7, fig.height = 6}
# Replicates Figure 3 of Jung 2022: 10 mg oral daily vs 114 mg (upper) and
# 146 mg (lower) weekly patches, median and 5th-95th percentiles.
rxode2::rxSetSeed(20220121)
fig3_times <- seq(0, 840, by = 2)
make_fig3 <- function(n, id_offset, dose_rows, label) {
  ids <- id_offset + seq_len(n)
  bind_rows(
    tidyr::crossing(id = ids, dose_rows) |> mutate(evid = 1L),
    tidyr::crossing(id = ids, time = fig3_times) |> mutate(amt = 0, cmt = "central", evid = 0L)
  ) |>
    mutate(regimen = label) |>
    arrange(id, time, desc(evid))
}
ev_fig3 <- bind_rows(
  make_fig3(100, 0L, tibble::tibble(time = 24 * (0:34), amt = 10, cmt = "depot_oral"), "Oral 10 mg daily"),
  make_fig3(100, 100L, tibble::tibble(time = 168 * (0:4), amt = 114, cmt = "depot_td"), "Patch 114 mg weekly"),
  make_fig3(100, 200L, tibble::tibble(time = 168 * (0:4), amt = 146, cmt = "depot_td"), "Patch 146 mg weekly")
)
stopifnot(!anyDuplicated(unique(ev_fig3[, c("id", "time", "evid")])))
sim3 <- rxode2::rxSolve(mod, events = ev_fig3, keep = "regimen", maxsteps = 1e6) |>
  as.data.frame()

summ3 <- sim3 |>
  group_by(regimen, time) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = quantile(Cc, 0.5), Q95 = quantile(Cc, 0.95),
            .groups = "drop")
oral3 <- filter(summ3, regimen == "Oral 10 mg daily")
bind_rows(
  bind_rows(oral3, filter(summ3, regimen == "Patch 114 mg weekly")) |> mutate(panel = "114 mg patch"),
  bind_rows(oral3, filter(summ3, regimen == "Patch 146 mg weekly")) |> mutate(panel = "146 mg patch")
) |>
  ggplot(aes(time, Q50, colour = regimen, fill = regimen)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line() +
  facet_wrap(~panel, ncol = 1) +
  scale_colour_manual(values = c("grey40", "forestgreen", "firebrick")) +
  scale_fill_manual(values = c("grey40", "forestgreen", "firebrick")) +
  labs(x = "Time (h)", y = "Donepezil (ng/mL)", colour = NULL, fill = NULL,
       caption = "Replicates Figure 3 of Jung 2022 (100 virtual subjects per regimen).") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Assumptions and deviations

- **Unrounded estimates from the later control stream.** Jung 2022 Table 3
  prints rounded estimates. Code S2 of Jung 2023 (the same authors' fractal
  kinetics paper) is the NONMEM control stream of this model, fitted to the same
  data with the identical OFV 1443.703; its `$THETA` and `$OMEGA` round to every
  Table 3 value and are used here.
- **The patch delivered fraction is fixed at the one-week wear.** Eq. 3 is a
  function of wear time, and Code S2 applies it to the patch dose amount in the
  dataset rather than inside the model. Every patch in the study and in the
  paper's simulations was worn for 168 h, so the model carries the resulting
  constant `F = 0.5513`. For a different wear time `D` (hours), override it with
  `ini(lfdepot_td = fixed(log(0.74 * 0.78257 * D / (D + 8.481))))`. The in-vitro
  dissolution data behind Eq. 3 (supplementary Table S3) are not needed to use
  the model; the fitted constants are printed in Eq. 3.
- **Patch removal is not modelled.** The drug that has entered the skin
  reservoir continues to transit to plasma after the patch comes off, as in Eq. 2
  and Code S2; the undelivered remainder is excluded by `F`.
- **Typographical error in Eq. 4.** The printed central-compartment equation
  has the oral input as `KA * SKIN`; Eq. 1 and Code S2 (`KA*A(1)`) both make it
  `KA * GUT`, which is what is implemented.
- **Bioequivalence simulation design.** The paper describes a 2 x 2 crossover
  with 100 simulated subjects per group but not how the variability between
  periods was generated (the model has no inter-occasion variability). The
  Table 4 comparison therefore uses the deterministic typical-subject ratio,
  which should be close to the published geometric-mean ratio, and checks it
  against the published 90% confidence intervals rather than recomputing them.
- **Study conduct.** The sponsor and analysts were in the Republic of Korea; the
  ethics approval was issued by the institutional review board of Raptim
  Research (Mumbai, India). The paper does not state where subjects were dosed.
- No erratum or correction notice for Jung 2022 was found as of 2026-09-30.
