---
title: "Benzathine benzylpenicillin G, subcutaneous infusion (Kado 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benzathine benzylpenicillin G, subcutaneous infusion (Kado 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Kado_2023_benzathine_benzylpenicillin_g"))$reference`
* Description: `r rxode2::rxode(readModelDb("Kado_2023_benzathine_benzylpenicillin_g"))$description`
* Article: <https://doi.org/10.1128/aac.00962-23>
* Supplement: `aac.00962-23-s0001.docx`, distributed with the open-access
  article (PMC10720493) and used here for the general covariate-model forms.

Benzathine penicillin G (BPG) has been the backbone of rheumatic heart disease
(RHD) secondary prophylaxis since 1955, given as a painful 1.2 MIU
intramuscular injection every four weeks. Kado 2023 is the phase 1
dose-escalation study of an alternative: a single high-dose *subcutaneous
infusion of penicillin* (SCIP) into the lower anterior abdomen, aiming for a
dosing interval of up to three months. This vignette reproduces the population
pharmacokinetic model built from that study.

The measured analyte is benzylpenicillin, whose own elimination half-life is
under an hour; everything observed over the 16-week follow-up is therefore
absorption-limited (flip-flop) kinetics driven by dissolution of the benzathine
salt at the infusion site. The paper handles this the same way the group's
earlier analyses did, by **fixing** the elimination rate constant and letting
the absorption model carry the shape of the curve. The absorption model that
won is unusual and is what makes this entry worth having in the library
(Figure 2 of the paper):

* the dose splits at the infusion site into a **bolus** portion and a
  **zero-order** portion, both entering the same depot;
* the zero-order portion is released over an estimated duration `DUR` of about
  44 days, which lengthens with increasing body mass index;
* depot contents pass through a single **transit** compartment
  (`t1/2,tr` = 0.702 h) and then a slow first-order absorption step
  (`t1/2,abs` = 11.8 days) into the central compartment.

# Population

Twenty-four healthy adult volunteers were recruited in Perth, Western
Australia, from 77 screened (Results, "Participants"). Four (16.7%) were
female. Median screening age was 26.9 years (range 18.0-54.1) and median body
mass index 25.1 kg/m^2 (range 21.9-34.0). Eight participants identified as
European, eight as Asian, four as Hispanic/Latino, two as African, and two as
of mixed race. Enrolment was deliberately balanced by body composition, with 12
participants in an "ideal" BMI stratum (20-24.9 kg/m^2) and 12 in a higher BMI
stratum (25-34.9 kg/m^2), because the effect of subcutaneous fat on absorption
was a secondary objective.

Participants received a single subcutaneous abdominal infusion, delivered over
up to 30 minutes under ultrasound guidance, in three escalating cohorts: 3.6
MIU (2,700 mg, n = 4), 7.2 MIU (5,400 mg, n = 10), and 10.8 MIU (8,100 mg,
n = 10). Dried blood spots were collected out to 112 days, giving 400
concentrations for the PK analysis; plasma and dried-blood-spot concentrations
were equivalent (bootstrap ratio 0.990, 95% CI 0.897-1.08), so the final model
was fitted to the dried-blood-spot data alone. The highest observed
benzylpenicillin concentration was below 300 ng/mL.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Kado_2023_benzathine_benzylpenicillin_g")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in
`inst/modeldb/specificDrugs/Kado_2023_benzathine_benzylpenicillin_g.R`. The
table below collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lkel` | 1.32 1/h per 70 kg, fixed (= 31.68 1/day) | Table 2, "k el (h -1 . 70 kg -1 )"; bootstrap column reads "Fixed" |
| `lvc` | 42.8 L per 70 kg | Table 2, "V/F (L . 70 kg -1 )" (bootstrap 42.8, 95% CI 40-46.1) |
| `e_wt_vc` | 1, fixed | Methods, "Population pharmacokinetic analysis": exponent of 1 for volume |
| `e_wt_cl` | 3/4, fixed | Methods, "Population pharmacokinetic analysis": exponent of 3/4 for clearance |
| `lthalf_abs` | 11.8 days | Table 2, "t 1/2, abs (days)" (bootstrap 11.8, 95% CI 10.8-12.8) |
| `lthalf_tr` | 0.702 h (= 0.02925 day) | Table 2, "t 1/2, tr (hours)" (bootstrap 0.69, 95% CI 0.519-0.873) |
| `ldur` | 1,062 h (= 44.25 day) | Table 2, "DUR (hours)" (bootstrap 1,055, 95% CI 832-1,250) |
| `lrat_bolus` | 2.54 | Table 2, "RATIO"; footnote b defines it as bolus over zero-order input |
| `e_bmi_dur` | 1.13 | Table 2, "Exponent of power BMI relationship on DUR" (RSE 28%) |
| BMI centering value | 25 kg/m^2 | Not printed; reconstructed (see Assumptions) from Results, "Pharmacokinetic modeling" |
| `etalvc` | omega^2 = 0.15^2 | Table 2, "IIV in V/F" = 15% (shrinkage 9%) |
| `etaldur` | omega^2 = 0.22^2 | Table 2, "IIV in DUR" = 22% (shrinkage 14%) |
| `etalthalf_abs` | omega^2 = 0.12^2 | Table 2, "IIV in t 1/2, abs" = 12% (shrinkage 26%) |
| `etalrat_bolus` | omega^2 = 0.88^2 | Table 2, "IIV in RAT" = 88% (shrinkage 16%) |
| `propSd` | 0.24 | Table 2, "RV (%)" = 24% (shrinkage 10%); Methods: additive error on log-transformed data |
| Dual-input depot, transit, `V/F`, `k el` topology | n/a | Figure 2, "Structure of the pharmacokinetic model" |
| Power covariate form `theta_i = theta_POP * (x_i / x_tilde)^theta_POWER` | n/a | Supplemental Information, "Population pharmacokinetic analysis" |
| IIV reported as `100% * sqrt(variability estimate)` | n/a | Table 2, footnote b |

The two conventions worth stating explicitly, because they are the usual place
a transcription goes wrong:

* **`RATIO` direction.** Table 2 footnote b defines RATIO as "the relative
  amount of drug assigned to bolus over zero-order input", so the zero-order
  fraction is `1 / (1 + 2.54)` = 28.2%. That reproduces the paper's own
  sentence that "the population estimate for RATIO corresponds to 28% of the
  dose passing through the zero-order absorption pathway", which fixes the
  direction unambiguously.
* **IIV scale.** Table 2 footnote b states that IIV and RV are reported as
  `100% * sqrt(variability estimate)`. The tabulated percentage is therefore
  `100 * omega` itself, so `omega^2 = (percentage / 100)^2` is used directly
  here -- *not* the `log(1 + CV^2)` conversion that applies when a paper
  reports a true log-normal CV%.

# Virtual cohort

Individual-level data from the trial are not public. Two virtual cohorts are
used below, both built with the correlation structure the paper specifies for
its own simulations ("1,000 simulated individuals (equal male and female) were
used with the correlation between weight and BMI set at 0.9"), scaled down to
the 200-per-arm cap used throughout this package.

```{r cohort}
set.seed(20231116)

n_per_arm <- 200L

# Bivariate log-normal (WT, BMI) with correlation 0.9 on the log scale.
# BMI is centred on the trial's median (25.1 kg/m^2) with a spread that
# approximately spans the observed 21.9-34.0 kg/m^2 range. Individual weights
# were never reported by the paper; the median is taken as
# BMI_median * (1.75 m)^2 = 77 kg. See "Assumptions and deviations".
make_cov <- function(n, id_offset = 0L) {
  rho <- 0.9
  z_wt  <- rnorm(n)
  z_bmi <- rho * z_wt + sqrt(1 - rho^2) * rnorm(n)
  tibble(
    id  = id_offset + seq_len(n),
    WT  = exp(log(77)   + 0.15 * z_wt),
    BMI = exp(log(25.1) + 0.12 * z_bmi)
  )
}

# Helper: dose rows + observation rows for one arm. The dose is written TWICE
# at each dosing time -- once as a bolus into `depot`, once into `depot2` with
# rate = -2 (modelled duration) so that dur(depot2) in the model delivers the
# zero-order fraction over DUR days. f() inside the model partitions the amount
# between the two paths, so both records carry the FULL dose amount.
make_arm <- function(cov_df, amt, dose_times, obs_times, label) {
  bind_rows(
    crossing(cov_df, time = dose_times) |>
      mutate(evid = 1L, amt = amt, cmt = "depot",  rate = 0),
    crossing(cov_df, time = dose_times) |>
      mutate(evid = 1L, amt = amt, cmt = "depot2", rate = -2),
    crossing(cov_df, time = obs_times) |>
      mutate(evid = 0L, amt = NA_real_, cmt = "central", rate = 0)
  ) |>
    mutate(arm = label) |>
    arrange(id, time, desc(evid))
}
```

The first cohort mirrors the trial itself: a single infusion in each of the
three dose groups, followed for 112 days.

```{r cohort-single}
dose_levels <- tibble::tribble(
  ~arm,       ~amt,
  "3.6 MIU",  2700,
  "7.2 MIU",  5400,
  "10.8 MIU", 8100
)

# Dense early grid: the transit compartment fills within a few hours
# (t1/2,tr = 0.702 h), so a uniform 0.5-day grid would place Tmax on the first
# post-dose sample and bias Cmax low.
obs_single <- sort(unique(c(seq(0, 2, by = 0.05), seq(0, 112, by = 0.5))))

ev_single <- bind_rows(lapply(seq_len(nrow(dose_levels)), function(i) {
  make_arm(
    cov_df     = make_cov(n_per_arm, id_offset = (i - 1L) * n_per_arm),
    amt        = dose_levels$amt[i],
    dose_times = 0,
    obs_times  = obs_single,
    label      = dose_levels$arm[i]
  )
}))
ev_single <- mutate(ev_single, arm = factor(arm, levels = dose_levels$arm))

# Disjoint IDs across arms are mandatory: rxSolve keys subjects on `id`, and a
# collision silently merges two subjects into one that receives both doses.
stopifnot(!anyDuplicated(unique(ev_single[, c("id", "time", "evid", "cmt")])))
stopifnot(n_distinct(ev_single$id) == n_per_arm * nrow(dose_levels))
```

The second cohort is the paper's dosing-interval scenario: 10.8 MIU repeated
every 13 weeks (91 days) for five doses, with concentrations recorded every 6
hours as in the paper's own simulations.

```{r cohort-ss}
tau      <- 91     # days, i.e. 13 weeks
n_doses  <- 5L
ev_ss <- make_arm(
  cov_df     = make_cov(n_per_arm, id_offset = 10000L),
  amt        = 8100,
  dose_times = tau * seq_len(n_doses) - tau,
  obs_times  = seq(0, tau * n_doses, by = 0.25),
  label      = "SCIP 10.8 MIU q13w"
)
```

# Simulation

```{r simulate}
mod <- readModelDb("Kado_2023_benzathine_benzylpenicillin_g")

sim_single <- rxode2::rxSolve(
  mod, events = ev_single, keep = c("arm", "WT", "BMI"), returnType = "data.frame"
)
sim_ss <- rxode2::rxSolve(
  mod, events = ev_ss, keep = c("WT", "BMI"), returnType = "data.frame"
)

# rxSolve has been observed to silently drop subjects; assert the count.
stopifnot(n_distinct(sim_single$id) == n_per_arm * nrow(dose_levels))
stopifnot(n_distinct(sim_ss$id) == n_per_arm)
```

A typical-value profile (all random effects zeroed) is used wherever the
published quantity is a point prediction rather than a distribution.

```{r simulate-typical}
mod_typical <- rxode2::zeroRe(mod)

ev_typical <- make_arm(
  cov_df     = tibble(id = 1L, WT = 70, BMI = 25),   # the model's reference individual
  amt        = 8100,
  dose_times = 0,
  obs_times  = seq(0, 112, by = 0.05),
  label      = "10.8 MIU typical"
)
sim_typical <- rxode2::rxSolve(
  mod_typical, events = ev_typical, omega = NA, sigma = NA, returnType = "data.frame"
) |>
  filter(!is.na(Cc))
```

## Checking the dual-input depot

The two-state depot is the one place this implementation departs from the
literal drawing in Figure 2, so it is checked explicitly: the bolus path must
receive exactly `RATIO / (RATIO + 1)` of the dose at time zero, and the
zero-order path must deliver the remainder at a constant rate that stops
exactly at `DUR`.

```{r depot-check}
ev_depot <- make_arm(
  cov_df     = tibble(id = 1L, WT = 70, BMI = 25),
  amt        = 8100,
  dose_times = 0,
  obs_times  = c(0, 10, 22, 44, 45, 60),
  label      = "depot check"
)
sim_depot <- rxode2::rxSolve(
  mod_typical, events = ev_depot, omega = NA, sigma = NA, returnType = "data.frame"
)

ratio    <- 2.54
durDays  <- 1062 / 24              # DUR at the reference BMI of 25 kg/m^2
ktr      <- log(2) / (0.702 / 24)  # transit rate constant, 1/day
rateZero <- 8100 / (ratio + 1) / durDays

# Bolus path: full bolus fraction present at t = 0.
stopifnot(isTRUE(all.equal(
  sim_depot$depot[sim_depot$time == 0], 8100 * ratio / (ratio + 1),
  tolerance = 1e-8
)))
# Zero-order path: while the infusion runs, depot2 sits at the steady state
# rate / ktr; after DUR it is empty.
stopifnot(isTRUE(all.equal(
  sim_depot$depot2[sim_depot$time %in% c(10, 22, 44)],
  rep(rateZero / ktr, 3), tolerance = 1e-4
)))
stopifnot(all(sim_depot$depot2[sim_depot$time > durDays] < 1e-6))

sim_depot |>
  select(time, depot, depot2, transit1, central, Cc) |>
  rename(
    "Time (day)"            = time,
    "depot (bolus, mg)"     = depot,
    "depot2 (zero-order, mg)" = depot2,
    "transit1 (mg)"         = transit1,
    "central (mg)"          = central,
    "Cc (ng/mL)"            = Cc
  ) |>
  knitr::kable(
    digits  = 3,
    caption = "Typical individual, 10.8 MIU. depot receives 8,100 * 2.54/3.54 = 5,811.9 mg as a bolus; depot2 holds the constant infusion steady state until DUR = 44.25 days and is empty thereafter."
  )
```

# Replicate published figures

## Figure 3 -- concentration-time profiles by dose cohort

Figure 3 of Kado 2023 is a prediction-corrected VPC of all 400 observations on
a log10 concentration scale, with the observed 10th, 50th, and 90th percentiles
inside their simulated 95% confidence bands. The panel below shows the
corresponding simulated percentiles from the packaged model for each dose
cohort.

```{r figure-3}
# Replicates Figure 3 of Kado 2023: simulated 10th / 50th / 90th percentiles of
# benzylpenicillin concentration over the 112-day follow-up, by dose cohort.
sim_single |>
  filter(!is.na(Cc), time > 0) |>
  group_by(arm, time) |>
  summarise(
    Q10 = quantile(Cc, 0.10),
    Q50 = quantile(Cc, 0.50),
    Q90 = quantile(Cc, 0.90),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.25) +
  geom_line() +
  geom_hline(yintercept = 20, linetype = "dashed", colour = "firebrick") +
  facet_wrap(~arm) +
  scale_y_log10() +
  labs(
    x = "Time (days)", y = "Benzylpenicillin (ng/mL)",
    title = "Figure 3 - simulated concentration percentiles by dose cohort",
    subtitle = "Median with 10th-90th percentile band; dashed line = 20 ng/mL PK/PD target",
    caption = "Replicates Figure 3 of Kado 2023."
  )
```

## Figure 4 -- 13-weekly high-dose SCIP against the PK/PD targets

Figure 4 of Kado 2023 compares 10.8 MIU SCIP given every 13 weeks against the
standard 1.2 MIU intramuscular dose every 4 weeks. Only the SCIP arm is
reproduced here -- the intramuscular comparator is not a model of *this* paper
(see "Assumptions and deviations").

```{r figure-4}
# Replicates the SCIP arm of Figure 4 of Kado 2023: median and 90% simulation
# interval over five 13-weekly 10.8 MIU subcutaneous infusions.
sim_ss |>
  filter(!is.na(Cc), time > 0) |>
  group_by(time) |>
  summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time / 7, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.3, fill = "grey30") +
  geom_line() +
  geom_hline(yintercept = c(10, 20), linetype = c("dotted", "dashed"),
             colour = c("steelblue", "firebrick")) +
  scale_y_log10() +
  labs(
    x = "Time (weeks)", y = "Benzylpenicillin (ng/mL)",
    title = "Figure 4 - SCIP 10.8 MIU every 13 weeks",
    subtitle = "Median with 90% simulation interval; 20 ng/mL (dashed) and 10 ng/mL (dotted) targets",
    caption = "Replicates the subcutaneous arm of Figure 4 of Kado 2023."
  )
```

## The BMI effect on the zero-order input duration

The paper states that the BMI power relationship moves `DUR` "over a range of
38-63 days for the lowest and highest BMI in the study". Reproducing that
statement is what identifies the unprinted centering value (see "Assumptions
and deviations").

```{r bmi-dur}
tibble(BMI = c(21.9, 25.1, 34.0)) |>
  mutate(
    label = c("lowest in study", "study median", "highest in study"),
    DUR_days = (1062 / 24) * (BMI / 25)^1.13
  ) |>
  rename("BMI (kg/m^2)" = BMI, "Participant" = label, "DUR (days)" = DUR_days) |>
  knitr::kable(
    digits  = 1,
    caption = "Zero-order input duration across the study BMI range. Kado 2023 reports this span as 38-63 days."
  )
```

# PKNCA validation

Kado 2023 reports no non-compartmental parameters, so PKNCA is used here as an
internal consistency check on the packaged model and as the route to the one
externally checkable exposure statement the paper does make (the highest
observed concentration). Because disposition is linear and every absorption
parameter is dose-independent, AUC and Cmax must scale exactly in proportion to
dose across the three cohorts; anything else would indicate a transcription
error in the dose-splitting or bioavailability terms.

```{r pknca}
# Keep only `!is.na(Cc)` -- adding `time > 0` or `Cc > 0` would drop the
# time-zero row that PKNCA needs to anchor AUC.
sim_nca <- sim_single |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, arm)

# Guarantee a time = 0 row per (id, arm); pre-dose extravascular Cc is 0.
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, arm) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, arm, time, .keep_all = TRUE) |>
  arrange(id, arm, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

dose_df <- ev_single |>
  filter(evid == 1, cmt == "depot") |>
  select(id, time, amt, arm)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)

intervals <- data.frame(
  start     = 0,
  end       = 112,
  cmax      = TRUE,
  tmax      = TRUE,
  auclast   = TRUE,
  half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r pknca-summary}
nca_tbl <- as.data.frame(nca_res) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "half.life")) |>
  group_by(arm, PPTESTCD) |>
  summarise(median = median(PPORRES), .groups = "drop") |>
  pivot_wider(names_from = PPTESTCD, values_from = median) |>
  mutate(
    dose_mg           = dose_levels$amt[match(as.character(arm), dose_levels$arm)],
    cmax_per_1000mg   = cmax / dose_mg * 1000,
    auclast_per_1000mg = auclast / dose_mg * 1000
  ) |>
  select(arm, cmax, tmax, auclast, half.life, cmax_per_1000mg, auclast_per_1000mg)

nca_tbl |>
  rename(
    "Dose cohort"               = arm,
    "Cmax (ng/mL)"              = cmax,
    "Tmax (day)"                = tmax,
    "AUC0-112d (ng*day/mL)"     = auclast,
    "t1/2 (day)"                = half.life,
    "Cmax per 1000 mg"          = cmax_per_1000mg,
    "AUC per 1000 mg"           = auclast_per_1000mg
  ) |>
  knitr::kable(
    digits  = c(0, 1, 2, 0, 2, 2, 1),
    caption = "Median simulated NCA parameters by dose cohort (n = 200 per arm). The dose-normalised columns agree to within a few percent; the residual spread is Monte-Carlo variation between independently drawn cohorts (see the typical-value check below)."
  )
```

The terminal half-life of roughly 12 days recovers `t1/2,abs` rather than the
sub-hour elimination half-life of benzylpenicillin -- the signature of the
flip-flop kinetics the paper describes. The dose-normalised columns agree across
the three cohorts to within a few percent; they are *not* exactly equal, because
each arm is an independently drawn cohort of 200 subjects, so the residual
spread is Monte-Carlo variation in the medians rather than nonlinearity. The
check below removes that noise and asserts exact proportionality.

```{r linearity}
# Exact dose-proportionality check. One typical individual (all random effects
# zeroed) at the reference covariates, run at each of the three dose levels: a
# linear model with dose-independent absorption must give exactly proportional
# Cmax and AUC. Any departure would indicate an error in the RATIO dose split,
# the f() partitioning, or dur(depot2).
ev_lin <- bind_rows(lapply(seq_len(nrow(dose_levels)), function(i) {
  make_arm(
    cov_df     = tibble(id = i, WT = 70, BMI = 25),
    amt        = dose_levels$amt[i],
    dose_times = 0,
    obs_times  = seq(0, 112, by = 0.05),
    label      = as.character(dose_levels$arm[i])
  )
}))

sim_lin <- rxode2::rxSolve(
  mod_typical, events = ev_lin, keep = "arm",
  omega = NA, sigma = NA, returnType = "data.frame"
) |>
  filter(!is.na(Cc))

trapz <- function(x, y) sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)

lin_tbl <- sim_lin |>
  arrange(arm, time) |>
  group_by(arm) |>
  summarise(
    Cmax = max(Cc),
    Tmax = time[which.max(Cc)],
    AUC  = trapz(time, Cc),
    .groups = "drop"
  ) |>
  mutate(
    dose_mg          = dose_levels$amt[match(arm, as.character(dose_levels$arm))],
    Cmax_per_1000mg  = Cmax / dose_mg * 1000,
    AUC_per_1000mg   = AUC  / dose_mg * 1000
  )

# Assert proportionality rather than eyeballing the table. The bound is a
# relative one because the ODE solver's absolute tolerance is not itself
# scale-invariant; 1e-5 is four orders of magnitude tighter than the
# Monte-Carlo spread seen in the stochastic table above.
relSpread <- function(x) diff(range(x)) / mean(x)
stopifnot(
  relSpread(lin_tbl$Cmax_per_1000mg) < 1e-5,
  relSpread(lin_tbl$AUC_per_1000mg)  < 1e-5,
  length(unique(lin_tbl$Tmax)) == 1L
)

lin_tbl |>
  select(arm, Cmax, Tmax, AUC, Cmax_per_1000mg, AUC_per_1000mg) |>
  rename(
    "Dose cohort"           = arm,
    "Cmax (ng/mL)"          = Cmax,
    "Tmax (day)"            = Tmax,
    "AUC0-112d (ng*day/mL)" = AUC,
    "Cmax per 1000 mg"      = Cmax_per_1000mg,
    "AUC per 1000 mg"       = AUC_per_1000mg
  ) |>
  knitr::kable(
    digits  = c(0, 1, 3, 0, 4, 3),
    caption = "Typical-value dose proportionality. The dose-normalised columns are exactly equal, as they must be for a linear, dose-independent model."
  )
```

`Tmax` sits at about 0.3 days (roughly 7 hours), well inside the first day: the
depot empties into the transit compartment with a half-life of 0.702 h, and the
central compartment tracks the transit outflow almost instantaneously because
benzylpenicillin's own elimination half-life is about half an hour. The
typical-value dose-normalised exposures (30.6 ng/mL and 735 ng*day/mL per
1,000 mg) run above the cohort medians in the previous table because the
reference individual weighs 70 kg while the virtual cohort's median weight is
77 kg, and both `V/F` and clearance scale up with weight.

# Comparison against published values

`nlmixr2lib::ncaComparisonTable()` is not used here because Kado 2023 publishes
no NCA table. The quantitative claims it *does* make are collected below and
compared against the packaged model.

```{r published-compare}
# Steady-state metrics over the final (fifth) 13-week dosing interval.
ss_last <- sim_ss |>
  filter(!is.na(Cc), time >= tau * (n_doses - 1), time <= tau * n_doses)

ss_metrics <- ss_last |>
  group_by(id) |>
  summarise(
    pct_above_20 = 100 * mean(Cc > 20),
    pct_above_10 = 100 * mean(Cc > 10),
    days_above_20 = tau * mean(Cc > 20),
    weeks_below_20 = tau * mean(Cc <= 20) / 7,
    .groups = "drop"
  )

med <- function(x) median(x)
si  <- function(x) sprintf("%.1f (%.1f-%.1f)", median(x),
                           quantile(x, 0.05), quantile(x, 0.95))

typical_peak     <- max(sim_typical$Cc)
zero_order_frac  <- 100 / (1 + 2.54)

comparison <- tibble::tribble(
  ~Quantity, ~Published, ~Simulated, ~Source,
  "Fraction of dose via the zero-order path (%)",
  "28", sprintf("%.1f", zero_order_frac),
  "Results, 'Pharmacokinetic modeling'",

  "Zero-order input duration DUR, population (days)",
  "~44", sprintf("%.1f", 1062 / 24),
  "Results, 'Pharmacokinetic modeling'",

  "DUR across the study BMI range (days)",
  "38-63", sprintf("%.0f-%.0f", (1062 / 24) * (21.9 / 25)^1.13,
                   (1062 / 24) * (34.0 / 25)^1.13),
  "Results, 'Pharmacokinetic modeling'",

  "Highest benzylpenicillin concentration, 10.8 MIU (ng/mL)",
  "< 300", sprintf("%.0f (typical individual)", typical_peak),
  "Results, 'Pharmacokinetic modeling'",

  "Time > 20 ng/mL, SCIP 10.8 MIU q13w (% of interval)",
  "63 (51-77)", si(ss_metrics$pct_above_20),
  "Results, 'Pharmacokinetic modeling'; Figure 4",

  "Time > 10 ng/mL, SCIP 10.8 MIU q13w (% of interval)",
  "75 (64-91)", si(ss_metrics$pct_above_10),
  "Results, 'Pharmacokinetic modeling'; Figure 4",

  "Time > 20 ng/mL, SCIP 10.8 MIU q13w (days)",
  "57", sprintf("%.0f", med(ss_metrics$days_above_20)),
  "Abstract",

  "Time <= 20 ng/mL per interval, SCIP (weeks)",
  "4.86", sprintf("%.2f", med(ss_metrics$weeks_below_20)),
  "Results, 'Pharmacokinetic modeling'"
)

knitr::kable(
  comparison,
  align   = c("l", "r", "r", "l"),
  caption = "Packaged model against every quantitative claim in Kado 2023. Published intervals are 90% simulation intervals; simulated values are median (5th-95th percentile) over 200 virtual subjects."
)
```

Every simulated value lands within 10% of the published figure -- the largest
gap is the time *below* 20 ng/mL per interval, 4.42 weeks against a published
4.86 (9%) -- so none of the threshold metrics is starred at the 20% tolerance
this package uses. The
remaining gap is in the expected direction: the paper simulated a US CDC
reference cohort of 20-year-olds, which is heavier than the virtual cohort used
here, and heavier individuals have larger `V/F` and clearance, hence lower
concentrations and slightly shorter time above target.

# Assumptions and deviations

* **BMI centering value reconstructed, not printed.** The supplement gives the
  power covariate form as `theta_i = theta_POP * (x_i / x_tilde)^theta_POWER`,
  where `x_tilde` is "the population average value of the covariate", but the
  numeric value of `x_tilde` for BMI appears nowhere in the paper or
  supplement. It is taken as **25 kg/m^2** here. Substituting the study's BMI
  range (21.9-34.0 kg/m^2) into the power form with `DUR` = 1,062 h and
  exponent 1.13 returns 38.1 and 62.6 days, which reproduces the paper's own
  statement that BMI moves `DUR` over "a range of 38-63 days for the lowest and
  highest BMI in the study"; the reported median BMI of 25.1 kg/m^2 returns
  62.4 days at the top of the range instead. 25 kg/m^2 is also the study's
  ideal-versus-higher BMI stratum boundary. The choice between 25.0 and 25.1
  changes `DUR` by under 0.5% and no conclusion in this vignette depends on it.
* **Single depot implemented as two states.** Figure 2 draws one depot fed by
  both a bolus and a zero-order input. rxode2 applies a single bioavailability
  per compartment, so the model carries `depot` (bolus path) and `depot2`
  (zero-order path) draining into the transit compartment with the *same* rate
  constant `ktr`. Because the depot outflow is first-order and the system is
  linear, `depot + depot2` obeys exactly the single-depot equation of Figure 2.
  This is an implementation device, not a structural change, but it means a
  user must write **two** dose records per administration -- a bolus into
  `depot` and a `rate = -2` record into `depot2`, both carrying the full dose
  amount, with `f()` doing the partitioning.
* **IIV interpreted as omega, not CV%.** Per Table 2 footnote b the tabulated
  IIV percentages are `100 * sqrt(variance)`, so `omega^2 = (pct/100)^2` is
  used directly. Note that the sibling model
  `Kado_2020_benzathine_benzylpenicillin_g` applies the alternative
  `omega^2 = log(1 + CV^2)` conversion to an identically worded footnote; the
  direct reading is used here because the footnote defines the quantity
  explicitly as the square root of the variance estimate. For the largest IIV
  in this model (88% on RATIO) the two readings differ materially, so this is
  recorded rather than left implicit.
* **Individual body weights were never reported.** The paper reports BMI but
  not weight, so the virtual cohort's weight distribution is an assumption:
  log-normal with median `25.1 * (1.75 m)^2` = 77 kg and a 15% log-scale
  spread, correlated with BMI at 0.9 on the log scale as the paper specifies
  for its own simulations. The BMI distribution is log-normal centred on the
  trial median with a 12% log-scale spread, approximately spanning the observed
  21.9-34.0 kg/m^2 range.
* **CDC reference cohort not reproduced.** The paper's simulations drew weight
  and BMI "from the CDC for 20-year-olds" with equal male and female
  representation. Those reference tables are not on disk, so the trial's own
  demographics were used instead rather than inventing percentiles. This is the
  main reason the simulated threshold metrics sit a few percent above the
  published ones.
* **Cohort size.** The paper simulated 1,000 individuals; this vignette uses
  200 per arm, the package cap.
* **Intramuscular comparator omitted from Figure 4.** The 1.2 MIU q4w
  intramuscular arm of Figure 4 is generated from the group's *earlier* model
  (reference 12), which is already packaged separately as
  `Kado_2020_benzathine_benzylpenicillin_g`. It is not a model of this paper,
  and that extraction carries a documented Cmax discrepancy of its own, so
  reproducing it here would import an unrelated uncertainty into this model's
  validation. Only the subcutaneous arm is shown.
* **Dosing-interval metrics evaluated at pseudo-steady state.** The published
  percentages describe repeated 13-weekly dosing, so metrics are computed over
  the fifth 91-day interval. Carryover between intervals is small (the
  typical-value profile is near 3-4 ng/mL at day 91), so the first and fifth
  intervals differ only slightly.
* **All parameter values come from the paper's own text, Table 2, Figure 2, and
  the open-access supplement.** No value was taken from an author
  communication, a digitised figure, or an upstream model file.
