---
title: "Oseltamivir (Kamal 2014)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Oseltamivir (Kamal 2014)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2)
suppressPackageStartupMessages({
  library(nlmixr2lib)
  library(PKNCA)
  library(rxode2)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Kamal_2014_oseltamivir"))$reference`
* Description: `r rxode2::rxode(readModelDb("Kamal_2014_oseltamivir"))$description`
* Article: <https://doi.org/10.1038/clpt.2014.120>

Oseltamivir is an inactive ethyl ester prodrug that is rapidly converted to
its active metabolite oseltamivir carboxylate (OC). Kamal 2014 pooled the
CASG114 (NCT00391768, United States) and WP22849 (NCT01286142, Europe)
prospective PK/PD studies to build a joint parent-plus-metabolite population
PK model in 133 influenza-infected infants under one year of age, and used it
to bridge infant exposures to doses already known to be safe and effective in
older children and adults. The resulting recommendation - 3 mg/kg twice daily
for all infants aged 2 weeks through 11 months - is the basis of the US Food
and Drug Administration label extension.

Structurally the model is a linear two-compartment model with first-order
absorption for the parent, feeding a one-compartment model for OC (three
disposition compartments in total). Because no OC was administered directly,
the OP-to-OC conversion fraction and the OC volume are not simultaneously
identifiable; the authors therefore assumed complete (100%) conversion and
estimated the OC central volume. The consequence for the implementation is
that the parent's elimination clearance `CL/F` *is* the OC formation
clearance: the entire parent central-compartment elimination flux enters
`central_oselcarb`.

# Population

The analysis data set comprised 604 oseltamivir and 648 OC plasma samples
from 133 infants (Kamal 2014 Results, first paragraph). Per Table 2 the
cohort had a mean body weight of 6.5 kg (SD 2.1, range 2.9-12.4), a mean
postnatal age of 23.5 weeks (SD 14.9, range 1.9-49.9), a mean gestational age
at birth of 38.1 weeks (SD 4.5, range 24.0-43.0), and a mean postconceptual
age of 61.6 weeks (SD 15.3, range 38.4-90.0). Of the subjects 59 (44.4%) were
female, and the race distribution was White 105 (78.9%), Black 14 (10.5%),
Other 12 (9.0%), Unknown 2 (1.5%). Enrolment by age band (Table 1) was
13 subjects (9.8%) aged 30 days or less, 33 (24.8%) at 31-90 days, 23 (17.3%)
at 91-180 days, 35 (26.3%) at 181-270 days, and 29 (21.8%) at 271 days or
more.

Dosing was weight-based, twice daily, and age-stratified: WP22849 gave
2 mg/kg (0-1 month), 2.5 mg/kg (1-3 months) and 3 mg/kg (3-12 months);
CASG114 gave 3 mg/kg (3-8 months) and 3 or 3.5 mg/kg (9 months and older).
Both trials excluded preterm neonates of postmenstrual age below 36 weeks.
Assay limits of quantitation were 1 ng/mL for oseltamivir and 50 ng/mL
for OC.

The same information is available programmatically via
`readModelDb("Kamal_2014_oseltamivir")()$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Kamal_2014_oseltamivir.R`. The
table below collects them in one place for review. Every fixed-effect,
variance and residual-error value comes from Kamal 2014 Table 3 (page 383),
"Estimate (95% CI)" column.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lka` (parent absorption `ka`) | 0.905 1/h | Table 3 theta1 (95% CI 0.691-1.12, 12.1% RSE) |
| `lcl` (parent `CL/F` at 8 kg) | 80.4 L/h | Table 3 theta2 (72.6-88.2, 4.96% RSE) |
| `lvc` (parent `V2/F` at 8 kg) | 165 L | Table 3 theta3 (124-209, 13.0% RSE) |
| `lq` (parent `Q/F` at 8 kg) | 19.6 L/h | Table 3 theta4 (17-22.1, 6.59% RSE) |
| `lvp` (parent `V3/F` at 8 kg) | 348 L | Table 3 theta5 (152-544, 28.7% RSE) |
| `lcl_oselcarb` (`CLM/F` at 8 kg, 24 wk) | 4.75 L/h | Table 3 theta6 (4.34-5.16, 4.40% RSE) |
| `lvc_oselcarb` (`VM/F` at 8 kg, 24 wk) | 40.2 L | Table 3 theta7 (36.6-43.8, 4.55% RSE) |
| `e_wt_cl` (allometric exponent, all CL terms) | 0.75, `fixed()` | Table 3 footnote; Methods "Allometric scaling and covariate model development" ("fixed power coefficients of 0.75 and 1") |
| `e_wt_vc` (allometric exponent, all V terms) | 1, `fixed()` | Table 3 footnote; same Methods paragraph |
| Reference weight | 8 kg | Table 3 footnote `(WT/8)`; Figure 2 caption ("the median body weight is 8 kg") |
| `e_age_cl_oselcarb` (`CLM,AGE` = `VM,AGE`) | 0.33 | Table 3 theta11 (0.265-0.396, 10.1% RSE) |
| Reference postnatal age | 24 weeks | Table 3 footnote `(AGE/24-1)`, "AGE is subject's age in weeks" |
| `Omega(1,1)` variance of `etalcl` | 0.154 | Table 3 (0.101-0.208, 17.6% RSE, CV 39.5%, shrinkage 7.2%) |
| `Omega(2,1)` covariance `etalcl`-`etalvc` | 0.157 | Table 3 (0.0576-0.256, R = 0.491) |
| `Omega(2,2)` variance of `etalvc` | 0.662 | Table 3 (0.294-1.03, CV 81.4%, shrinkage 13.3%) |
| `Omega(3,1)` covariance `etalcl`-`etalcl_oselcarb` | 0.0638 | Table 3 (0.0284-0.0991, R = 0.455) |
| `Omega(3,2)` covariance `etalvc`-`etalcl_oselcarb` | 0.169 | Table 3 (0.0875-0.251, R = 0.584) |
| `Omega(3,3)` variance of `etalcl_oselcarb` | 0.127 | Table 3 (0.1-0.154, CV 35.7%, shrinkage 2.7%) |
| `propSd` (parent proportional residual SD) | 0.467 | Table 3 theta8 (0.420-0.515, CV 46.7%) |
| `propSd_oselcarb` (OC proportional residual SD) | 0.118 | Table 3 theta9 (0.094-0.141, CV 11.8%) |
| `addSd_oselcarb` (OC additive residual SD) | 39.5 ng/mL = 0.0395 mg/L | Table 3 theta10 (32.9-46.1, 8.53% RSE) |
| ODE `d/dt(depot) = -ka * depot` | n/a | Methods "Population pharmacokinetic analysis" ("First-order absorption ... were assumed") |
| ODE `d/dt(central) = ka*depot - kel*central - k12*central + k21*peripheral1` | n/a | Methods ("oseltamivir was described by a linear two-compartment model with first-order absorption") |
| ODE `d/dt(peripheral1) = k12*central - k21*peripheral1` | n/a | same |
| ODE `d/dt(central_oselcarb) = kel*central - kel_oselcarb*central_oselcarb` | n/a | Methods ("direct first-order conversion of parent drug to metabolite"; "complete (100%) oseltamivir to OC metabolism was assumed") |
| `CLM/F ~ 1 + CLM,AGE*(AGE/24 - 1)` | n/a | Table 3 footnote |
| `VM/F ~ 1 + VM,AGE*(AGE/24 - 1)`, `VM,AGE = CLM,AGE` | n/a | Table 3 footnote |

The three correlations printed in Table 3's "Variability" column reproduce
exactly from the tabulated covariances using the footnote's own definition
`R = Omega(2,1)/sqrt(Omega(1,1)*Omega(2,2))`, which is what establishes that
the Omega entries are variances rather than SDs or %CVs:

```{r omega-check}
om <- matrix(c(0.154,  0.157,  0.0638,
               0.157,  0.662,  0.169,
               0.0638, 0.169,  0.127), nrow = 3, byrow = TRUE)
r <- stats::cov2cor(om)
omega_check <- tibble::tibble(
  pair      = c("CL-V2", "CL-CLM", "V2-CLM"),
  printed_R = c(0.491, 0.455, 0.584),
  derived_R = c(r[2, 1], r[3, 1], r[3, 2])
)
omega_check |>
  dplyr::rename("Eta pair" = pair,
                "R printed in Table 3" = printed_R,
                "R derived from Omega" = derived_R) |>
  knitr::kable(digits = 4,
               caption = "Table 3 correlations reproduced from the tabulated Omega entries.")

# The block must also be positive definite for rxode2's Cholesky sampler.
stopifnot(
  all(abs(omega_check$printed_R - omega_check$derived_R) < 0.002),
  all(eigen(om, only.values = TRUE)$values > 0)
)
```

# Virtual cohort

The original subject-level data are not publicly available, so the cohort
below is synthetic. Postnatal ages are drawn from the five enrolment bands of
Kamal 2014 Table 1 using the published band proportions, and body weight is
drawn from a weight-for-age relationship calibrated so that the simulated
cohort reproduces the Table 2 marginal weight statistics (mean 6.5 kg,
SD 2.1, range 2.9-12.4). The calibration targets are asserted below, so the
cohort construction is a gate rather than a decoration.

```{r cohort}
rxode2::rxSetSeed(20140120)
set.seed(20140120)

n_subj <- 200L

# Kamal 2014 Table 1 enrolment bands (days) and their published proportions.
# The band limits are clipped to the Table 2 observed age range 1.9-49.9
# weeks (13-349 days).
age_bands <- tibble::tibble(
  lo_day = c(13,  31,  91, 181, 271),
  hi_day = c(30,  90, 180, 270, 349),
  n_pub  = c(13L, 33L, 23L, 35L, 29L)
)
stopifnot(sum(age_bands$n_pub) == 133L)

n_band <- diff(c(0, round(cumsum(age_bands$n_pub / sum(age_bands$n_pub)) * n_subj)))
age_day <- unlist(Map(stats::runif, n_band, age_bands$lo_day, age_bands$hi_day))

# Weight-for-age. Median weight rises roughly linearly across the first year
# in this cohort; the slope/intercept and the 13% lognormal residual are
# chosen so the simulated marginal weight distribution matches Table 2.
# See "Assumptions and deviations".
age_week <- age_day / 7
wt <- (3.2 + 0.14 * age_week) * exp(stats::rnorm(n_subj, 0, 0.13))
wt <- pmin(pmax(wt, 2.9), 12.4)

cohort <- tibble::tibble(
  id       = seq_len(n_subj),
  AGE_WEEK = age_week,
  PNA      = age_day / 30.4375,     # canonical PNA is postnatal age in MONTHS
  WT       = wt
) |>
  dplyr::mutate(
    age_group = cut(
      PNA, breaks = c(-Inf, 1, 3, 6, 9, Inf),
      labels = c("0-1 mo", "1-3 mo", "3-6 mo", "6-9 mo", "9-12 mo"),
      right = FALSE
    )
  )

cohort_summary <- tibble::tibble(
  Statistic  = c("Mean weight (kg)", "SD weight (kg)", "Min weight (kg)", "Max weight (kg)",
                 "Mean age (weeks)", "SD age (weeks)", "Min age (weeks)", "Max age (weeks)"),
  Published  = c(6.5, 2.1, 2.9, 12.4, 23.5, 14.9, 1.9, 49.9),
  Simulated  = c(mean(cohort$WT), stats::sd(cohort$WT), min(cohort$WT), max(cohort$WT),
                 mean(cohort$AGE_WEEK), stats::sd(cohort$AGE_WEEK),
                 min(cohort$AGE_WEEK), max(cohort$AGE_WEEK))
)
knitr::kable(cohort_summary, digits = 2,
             caption = "Simulated virtual cohort versus the Kamal 2014 Table 2 marginal statistics.")

# Gate the cohort construction on the CENTRE of each distribution, which is
# reproducible; the min/max are single order statistics of a random draw and
# are reported but not asserted.
stopifnot(
  abs(mean(cohort$WT) - 6.5) < 0.5,
  abs(stats::sd(cohort$WT) - 2.1) < 0.5,
  abs(mean(cohort$AGE_WEEK) - 23.5) < 2.5,
  abs(stats::sd(cohort$AGE_WEEK) - 14.9) < 2.5
)
```

# Structural gate: closed-form exposure identities

Before any cohort simulation, check the packaged model against two identities
that follow directly from its structure and hold exactly for a linear model
at steady state:

* parent `AUC(0-tau),ss = Dose / (CL/F)`, and
* OC `AUC(0-tau),ss = Dose / (CLM/F)` - which holds *only* because the
  authors assumed complete conversion, so the mass balance routes the whole
  parent elimination flux into the OC compartment.

These are pure numerical checks on a typical-value (zero random effect)
subject, so a tight tolerance is appropriate. The regimen is extended to
ten days here purely so that steady state is fully attained; the parent's
terminal half-life is about 15 h, so the five-day regimen the paper
simulated is at roughly 99.4% of steady state and would understate both
AUCs by a few tenths of a percent.

```{r structural-gate}
mod         <- rxode2::rxode(readModelDb("Kamal_2014_oseltamivir"))
mod_typical <- rxode2::zeroRe(mod)

ref_wt  <- 8
ref_pna <- 24 * 7 / 30.4375   # 24 weeks expressed in months

typ_wt   <- 6.5
typ_week <- 23.5
typ_pna  <- typ_week * 7 / 30.4375

make_events <- function(subjects, mgkg, n_dose, tau = 12,
                        dense_from = NULL, dense_by = 0.25, coarse_by = 4) {
  last <- (n_dose - 1) * tau
  if (is.null(dense_from)) dense_from <- last
  dose_times <- seq(0, last, by = tau)
  obs_times  <- sort(unique(c(
    seq(0, dense_from, by = coarse_by),
    seq(dense_from, dense_from + tau, by = dense_by)
  )))
  doses <- subjects |>
    tidyr::expand_grid(time = dose_times) |>
    dplyr::mutate(evid = 1L, amt = mgkg * WT, cmt = "depot",
                  dvid = NA_integer_)
  obs <- subjects |>
    tidyr::expand_grid(time = obs_times) |>
    dplyr::mutate(evid = 0L, amt = 0, cmt = "central", dvid = 1L)
  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

trap <- function(x, y) sum(diff(x) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)

typ_subject <- tibble::tibble(id = 1L, WT = typ_wt, PNA = typ_pna)
ev_typ <- make_events(typ_subject, mgkg = 3, n_dose = 20)
stopifnot(!anyDuplicated(ev_typ[, c("id", "time", "evid")]))

sim_typ <- rxode2::rxSolve(mod_typical, events = ev_typ,
                           keep = c("WT", "PNA"), useLinCmt = FALSE) |>
  as.data.frame()

tau_start <- 19 * 12
ss <- sim_typ |>
  dplyr::filter(time >= tau_start, time <= tau_start + 12) |>
  dplyr::distinct(time, .keep_all = TRUE)

amt_typ <- 3 * typ_wt
identity_check <- tibble::tibble(
  Analyte    = c("Oseltamivir", "Oseltamivir carboxylate"),
  `Closed form`  = c(amt_typ / unique(ss$cl), amt_typ / unique(ss$cl_oselcarb)),
  `Simulated`    = c(trap(ss$time, ss$Cc), trap(ss$time, ss$Cc_oselcarb))
) |>
  dplyr::mutate(Ratio = Simulated / `Closed form`)

knitr::kable(identity_check, digits = 5,
             caption = "Steady-state AUC(0-12) identities for a typical 6.5 kg, 23.5-week infant on 3 mg/kg twice daily (mg*h/L).")

stopifnot(all(abs(identity_check$Ratio - 1) < 0.005))
```

The individual parameters themselves must also reproduce the published
covariate equations exactly:

```{r param-gate}
allom_cl <- (typ_wt / ref_wt)^0.75
allom_v  <- (typ_wt / ref_wt)^1
age_fac  <- 1 + 0.33 * (typ_pna / ref_pna - 1)

param_check <- tibble::tibble(
  Parameter = c("CL/F (L/h)", "V2/F (L)", "Q/F (L/h)", "V3/F (L)",
                "CLM/F (L/h)", "VM/F (L)"),
  `Published equation` = c(80.4 * allom_cl, 165 * allom_v, 19.6 * allom_cl,
                           348 * allom_v, 4.75 * allom_cl * age_fac,
                           40.2 * allom_v * age_fac),
  Model = c(unique(ss$cl), unique(ss$vc), unique(ss$q), unique(ss$vp),
            unique(ss$cl_oselcarb), unique(ss$vc_oselcarb))
)
knitr::kable(param_check, digits = 4,
             caption = "Typical-value parameters versus the Kamal 2014 Table 3 footnote equations.")
stopifnot(all(abs(param_check$Model / param_check$`Published equation` - 1) < 1e-8))
```

The paper summarises the age effect as OC clearance and volume "increasing
linearly with age (~72% per year)". That figure follows from the single
estimated coefficient:

```{r age-slope-gate}
per_year <- 0.33 * (365.25 / 7) / 24
stopifnot(abs(per_year - 0.72) < 0.01)
per_year
```

# Replicate published figures

## Figure 2 - allometrically normalised OC clearance versus age

Kamal 2014 Figure 2 plots OC clearance normalised for allometrically scaled
body weight against infant age, and the Discussion states the plot "confirms
the effect of maturation estimated by the model". The normalisation divides
the individual `CLM/F` by `(WT/8)^0.75`, which leaves the linear age function
and the OC clearance random effect. Simulating the cohort first gives both
the individual parameters and the concentration profiles used below.

```{r simulate}
rxode2::rxSetSeed(20140121)
set.seed(20140121)

ev_3 <- make_events(cohort |> dplyr::select(id, WT, PNA, age_group),
                    mgkg = 3, n_dose = 10)
stopifnot(!anyDuplicated(ev_3[, c("id", "time", "evid")]))

sim_3 <- rxode2::rxSolve(mod, events = ev_3,
                         keep = c("WT", "PNA", "age_group"),
                         useLinCmt = FALSE) |>
  as.data.frame()
```

```{r figure-2, fig.cap = "Replicates Figure 2 of Kamal 2014: OC clearance normalised by allometrically scaled body weight versus postnatal age."}
per_subject <- sim_3 |>
  dplyr::distinct(id, .keep_all = TRUE) |>
  dplyr::select(id, WT, PNA, age_group, cl_oselcarb, vc_oselcarb, cl, vc) |>
  dplyr::mutate(AGE_WEEK = PNA * 30.4375 / 7,
                clm_norm = cl_oselcarb / (WT / ref_wt)^0.75)

typical_line <- tibble::tibble(AGE_WEEK = seq(0, 52, by = 1)) |>
  dplyr::mutate(clm_norm = 4.75 * (1 + 0.33 * (AGE_WEEK / 24 - 1)))

ggplot(per_subject, aes(AGE_WEEK, clm_norm)) +
  geom_point(alpha = 0.45, size = 1.3, colour = "grey30") +
  geom_line(data = typical_line, colour = "firebrick", linewidth = 0.9) +
  scale_x_continuous(name = "Postnatal age (weeks)", limits = c(0, 52)) +
  scale_y_continuous(name = "CLM/F normalised to 8 kg (L/h)") +
  labs(title = "Figure 2: allometrically normalised OC clearance versus age",
       subtitle = "Points = simulated individuals; red line = typical value 4.75 * (1 + 0.33 * (AGE/24 - 1))") +
  theme_minimal()
```

A regression of the normalised clearance on age must recover the published
coefficient, because the normalisation removes the only other covariate:

```{r figure-2-gate}
fit <- stats::lm(clm_norm ~ I(AGE_WEEK / 24 - 1), data = per_subject)
recovered <- stats::coef(fit)
tibble::tibble(
  Quantity  = c("Intercept: CLM/F at 24 weeks, 8 kg (L/h)",
                "Slope: CLM/F per unit (AGE/24 - 1) (L/h)",
                "Implied CLM,AGE coefficient"),
  Published = c(4.75, 4.75 * 0.33, 0.33),
  Recovered = c(recovered[[1]], recovered[[2]], recovered[[2]] / recovered[[1]])
) |>
  knitr::kable(digits = 4,
               caption = "Least-squares recovery of the Kamal 2014 age effect from the simulated cohort.")

# The intercept and the implied coefficient are cohort means, so they are
# reproducible; a residual-error-free regression on 200 subjects with 35.7%
# CV on CLM recovers them to within a few percent.
stopifnot(
  abs(recovered[[1]] / 4.75 - 1) < 0.10,
  abs(recovered[[2]] / recovered[[1]] / 0.33 - 1) < 0.20
)
```

## Figure 4 - steady-state OC exposure by age subgroup at 3 mg/kg

Kamal 2014 Figure 4 shows box-and-whisker distributions of steady-state OC
AUC, Cmin and Cmax by age subgroup (0-1, 1-3, 3-6, 6-9 and 9-12 months) for
a 3 mg/kg twice-daily regimen simulated over five days. The corresponding
simulated distributions are below.

```{r figure-4-exposure}
ss_3 <- sim_3 |>
  dplyr::filter(time >= 108, time <= 120) |>
  dplyr::distinct(id, time, .keep_all = TRUE) |>
  dplyr::mutate(t_in_interval = time - 108)

exposure <- ss_3 |>
  dplyr::group_by(id, age_group, WT, PNA) |>
  dplyr::summarise(
    AUC_oc  = trap(t_in_interval, Cc_oselcarb) * 1000,   # mg*h/L -> h*ng/mL
    Cmax_oc = max(Cc_oselcarb) * 1000,
    Cmin_oc = min(Cc_oselcarb) * 1000,
    AUC_op  = trap(t_in_interval, Cc) * 1000,
    Cmax_op = max(Cc) * 1000,
    .groups = "drop"
  )
stopifnot(nrow(exposure) == n_subj, all(is.finite(exposure$AUC_oc)))
```

The same cohort is also solved with the random effects zeroed, which gives the
typical-value exposure implied by each subject's covariates alone. That
typical-value series is what the structural claims below are asserted
against; see the discussion under the gate for why the stochastic group
medians are not a reproducible statistic at these group sizes.

```{r figure-4-typical}
sim_3_typ <- rxode2::rxSolve(mod_typical, events = ev_3,
                             keep = c("WT", "PNA", "age_group"),
                             useLinCmt = FALSE) |>
  as.data.frame()

exposure_typ <- sim_3_typ |>
  dplyr::filter(time >= 108, time <= 120) |>
  dplyr::distinct(id, time, .keep_all = TRUE) |>
  dplyr::mutate(t_in_interval = time - 108) |>
  dplyr::group_by(id, age_group) |>
  dplyr::summarise(AUC_oc = trap(t_in_interval, Cc_oselcarb) * 1000,
                   .groups = "drop")
stopifnot(nrow(exposure_typ) == n_subj)
```

```{r figure-4, fig.height = 3.6, fig.cap = "Replicates Figure 4 of Kamal 2014: steady-state OC AUC(0-12), Cmin and Cmax by age subgroup after 3 mg/kg twice daily."}
exposure |>
  dplyr::select(age_group, AUC_oc, Cmin_oc, Cmax_oc) |>
  tidyr::pivot_longer(-age_group, names_to = "metric", values_to = "value") |>
  dplyr::mutate(metric = factor(
    metric, levels = c("AUC_oc", "Cmin_oc", "Cmax_oc"),
    labels = c("AUC(0-12) (h*ng/mL)", "Cmin (ng/mL)", "Cmax (ng/mL)"))) |>
  ggplot(aes(age_group, value)) +
  geom_boxplot(outlier.size = 0.7, fill = "grey90") +
  facet_wrap(~metric, scales = "free_y") +
  labs(x = "Age subgroup", y = NULL,
       title = "Figure 4: steady-state OC exposure by age subgroup, 3 mg/kg twice daily") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

```{r figure-4-table}
exposure_by_group <- exposure |>
  dplyr::group_by(age_group) |>
  dplyr::summarise(
    n            = dplyr::n(),
    median_AUC   = stats::median(AUC_oc),
    q25_AUC      = stats::quantile(AUC_oc, 0.25),
    q75_AUC      = stats::quantile(AUC_oc, 0.75),
    median_Cmin  = stats::median(Cmin_oc),
    median_Cmax  = stats::median(Cmax_oc),
    .groups = "drop"
  )

exposure_by_group |>
  dplyr::rename("Age subgroup" = age_group,
                "N" = n,
                "Median AUC(0-12) (h*ng/mL)" = median_AUC,
                "Q1 AUC" = q25_AUC,
                "Q3 AUC" = q75_AUC,
                "Median Cmin (ng/mL)" = median_Cmin,
                "Median Cmax (ng/mL)" = median_Cmax) |>
  knitr::kable(digits = 0,
               caption = "Simulated steady-state OC exposure by age subgroup at 3 mg/kg twice daily.")
```

Kamal 2014 reports that "the 3 mg/kg dose produced higher exposures in
neonates (aged 0-1 month)" and that there was "a modest increase in median
AUC for OC in the youngest infants". That is a structural consequence of the
age effect - `CLM/F` rises with age, so weight-normalised exposure falls -
and is therefore a monotone claim the packaged model must reproduce. It is
asserted on the typical-value series:

```{r figure-4-gate}
med_typ <- exposure_typ |>
  dplyr::group_by(age_group) |>
  dplyr::summarise(median_AUC = stats::median(AUC_oc), .groups = "drop")
med_t <- stats::setNames(med_typ$median_AUC, as.character(med_typ$age_group))
med_s <- stats::setNames(exposure_by_group$median_AUC,
                         as.character(exposure_by_group$age_group))

tibble::tibble(
  `Age subgroup` = names(med_t),
  `Typical-value median AUC (h*ng/mL)` = as.numeric(med_t),
  `Cohort median AUC (h*ng/mL)` = as.numeric(med_s)
) |>
  knitr::kable(digits = 0,
               caption = "Median steady-state OC AUC(0-12) by age subgroup, typical-value versus stochastic cohort.")

stopifnot(
  # Structural claim: with the random effects zeroed, median OC AUC falls
  # monotonically across the five age subgroups, and the youngest subgroup
  # exceeds the oldest by about 1.4-fold. The covariates are drawn with base
  # R's RNG under a fixed seed, so this series is exactly reproducible.
  all(diff(med_t) < 0),
  med_t[[1]] / med_t[[length(med_t)]] > 1.35,
  # Stochastic cohort: DIRECTION only. Adjacent subgroups differ by 6-10% in
  # typical exposure, far below the 35.7% CV per-subject spread on CLM/F, so
  # with 20-53 subjects per group the adjacent-pair ordering of the sample
  # medians is sampling noise, not a model property (monotone in only ~40%
  # of eta draws). Only the youngest-versus-oldest contrast, which spans the
  # full 1.4-fold effect, is stable enough to assert.
  med_s[[1]] > med_s[[length(med_s)]]
)
```

The typical-value series is strictly monotone and spans
`r sprintf("%.2f", med_t[[1]] / med_t[[length(med_t)]])`-fold, reproducing the
paper's "modest increase in median AUC for OC in the youngest infants". The
stochastic cohort medians in the same table are noticeably rougher, which is
the expected consequence of a 35.7% CV random effect on `CLM/F` acting on
subgroups of 20-53 subjects.

## Dose linearity across the tested 2 to 3.5 mg/kg range

The paper simulated 2, 2.5, 3 and 3.5 mg/kg regimens and reported that "the
lower doses of 2.5 and 2 mg/kg showed a tendency toward lower exposures than
benchmark comparators". The model is linear in dose, so exposure must scale
exactly with the mg/kg level; that exactness is itself a check that nothing
in the covariate or absorption code introduces an unintended nonlinearity.

```{r dose-linearity}
dose_levels <- c(2, 2.5, 3, 3.5)
dose_sim <- lapply(dose_levels, function(d) {
  ev <- make_events(typ_subject, mgkg = d, n_dose = 20)
  s  <- rxode2::rxSolve(mod_typical, events = ev, keep = c("WT", "PNA"),
                        useLinCmt = FALSE) |>
    as.data.frame() |>
    dplyr::filter(time >= tau_start, time <= tau_start + 12) |>
    dplyr::distinct(time, .keep_all = TRUE)
  tibble::tibble(mgkg = d,
                 AUC_oc  = trap(s$time - tau_start, s$Cc_oselcarb) * 1000,
                 Cmax_oc = max(s$Cc_oselcarb) * 1000,
                 Cmin_oc = min(s$Cc_oselcarb) * 1000)
}) |>
  dplyr::bind_rows() |>
  dplyr::mutate(AUC_ratio_vs_3 = AUC_oc / AUC_oc[mgkg == 3])

dose_sim |>
  dplyr::rename("Dose (mg/kg BID)" = mgkg,
                "OC AUC(0-12) (h*ng/mL)" = AUC_oc,
                "OC Cmax (ng/mL)" = Cmax_oc,
                "OC Cmin (ng/mL)" = Cmin_oc,
                "AUC ratio vs 3 mg/kg" = AUC_ratio_vs_3) |>
  knitr::kable(digits = c(1, 0, 0, 0, 4),
               caption = "Typical-subject steady-state OC exposure across the doses simulated in Kamal 2014.")

stopifnot(all(abs(dose_sim$AUC_ratio_vs_3 - dose_sim$mgkg / 3) < 1e-6))
```

# PKNCA validation

NCA is run over the steady-state 12-hour interval, once per analyte, with the
age subgroup as the treatment grouping variable so per-group results line up
with Figure 4. The interval already contains its own start record (the
steady-state trough at the shifted time zero), so no synthetic time-zero row
is required and no "AUC range starting before the first measurement" warning
arises.

```{r pknca-setup}
nca_dose <- exposure |>
  dplyr::transmute(id, age_group, time = 0, amt = 3 * WT)

run_nca <- function(conc_df) {
  conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | age_group + id)
  dose_obj <- PKNCA::PKNCAdose(nca_dose, amt ~ time | age_group + id)
  intervals <- data.frame(start = 0, end = 12,
                          cmax = TRUE, tmax = TRUE, cmin = TRUE,
                          auclast = TRUE)
  # Pure linear trapezoid, so the PKNCA result is directly comparable with
  # the trapezoidal integral computed above and with the closed-form
  # Dose/CLM identity. PKNCA's default is linear-up / log-down, which is the
  # right choice for sparse observed data but would introduce a method
  # difference into what is meant to be an exact cross-check.
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals,
                                 options = list(auc.method = "linear")))
}

conc_oc <- ss_3 |>
  dplyr::transmute(id, age_group, time = t_in_interval, Cc = Cc_oselcarb) |>
  dplyr::filter(!is.na(Cc))
conc_op <- ss_3 |>
  dplyr::transmute(id, age_group, time = t_in_interval, Cc = Cc) |>
  dplyr::filter(!is.na(Cc))
stopifnot(nrow(conc_oc) > 0, nrow(conc_op) > 0,
          all(c(0, 12) %in% unique(conc_oc$time)))

nca_oc <- run_nca(conc_oc)
nca_op <- run_nca(conc_op)
```

```{r pknca-tables}
summarise_nca <- function(res, label) {
  as.data.frame(res$result) |>
    dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "cmin", "auclast")) |>
    dplyr::group_by(age_group, PPTESTCD) |>
    dplyr::summarise(value = stats::median(PPORRES), .groups = "drop") |>
    tidyr::pivot_wider(names_from = PPTESTCD, values_from = value) |>
    dplyr::mutate(analyte = label)
}

dplyr::bind_rows(summarise_nca(nca_oc, "Oseltamivir carboxylate"),
                 summarise_nca(nca_op, "Oseltamivir")) |>
  dplyr::relocate(analyte) |>
  dplyr::rename("Analyte" = analyte,
                "Age subgroup" = age_group,
                "Cmax (mg/L)" = cmax,
                "Cmin (mg/L)" = cmin,
                "Tmax (h)" = tmax,
                "AUClast 0-12 (mg*h/L)" = auclast) |>
  knitr::kable(digits = 3,
               caption = "Median PKNCA parameters over the steady-state 12-hour interval, 3 mg/kg twice daily.")
```

Half-life is deliberately not requested: a steady-state dosing interval
contains no post-absorption washout, so a terminal slope fitted inside it
would not estimate the disposition half-life of either analyte.

The PKNCA AUC must agree with the trapezoidal exposure computed above and,
for OC, with the `Dose/CLM` identity applied per subject:

```{r pknca-gate}
auc_pknca <- as.data.frame(nca_oc$result) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::transmute(id, auc_pknca = PPORRES * 1000)

auc_cmp <- exposure |>
  dplyr::select(id, WT, AUC_oc) |>
  dplyr::left_join(auc_pknca, by = "id") |>
  dplyr::left_join(per_subject |> dplyr::select(id, cl_oselcarb), by = "id") |>
  dplyr::mutate(auc_closed = 3 * WT / cl_oselcarb * 1000)
stopifnot(nrow(auc_cmp) == n_subj, !anyNA(auc_cmp$auc_pknca))

tibble::tibble(
  Comparison = c("PKNCA auclast vs trapezoidal AUC",
                 "PKNCA auclast vs per-subject Dose/CLM"),
  `Median ratio` = c(stats::median(auc_cmp$auc_pknca / auc_cmp$AUC_oc),
                     stats::median(auc_cmp$auc_pknca / auc_cmp$auc_closed)),
  `Max |deviation|` = c(max(abs(auc_cmp$auc_pknca / auc_cmp$AUC_oc - 1)),
                        max(abs(auc_cmp$auc_pknca / auc_cmp$auc_closed - 1)))
) |>
  knitr::kable(digits = 5,
               caption = "PKNCA cross-checks against the trapezoidal integral and the closed-form steady-state identity.")

stopifnot(
  # Same integral, same grid: agreement is exact to numerical precision.
  max(abs(auc_cmp$auc_pknca / auc_cmp$AUC_oc - 1)) < 1e-8,
  # Against the closed form, the five-day regimen sits at ~99.4% of steady
  # state, so a sub-1% shortfall is expected and is asserted as such.
  abs(stats::median(auc_cmp$auc_pknca / auc_cmp$auc_closed) - 1) < 0.01
)
```

# Comparison against the published bridging criteria

Kamal 2014 reports no NCA table for the infant cohort itself - Figure 4 is
graphical only - so the quantitative published targets are the pharmacokinetic
bridging criteria. These were derived from a study of children aged 1-2 years
given a single 30 mg dose, in which "mean total OC AUC (AUCinf) was
3,905 h*ng/ml, mean minus 1 SD AUCinf was 2,618 h*ng/ml, and the minimum
AUCinf was 1,807 h*ng/ml". The paper then required that the simulated infant
steady-state AUC(0-12) exceed 3,905 in 50% of subjects, 2,618 in 84%, and
1,807 in 95%, and concluded that 3 mg/kg twice daily satisfied all three.

```{r bridging}
auc_all <- exposure$AUC_oc
bridging <- tibble::tibble(
  Criterion = c("AUC(0-12) > 3,905 h*ng/mL in 50% of subjects",
                "AUC(0-12) > 2,618 h*ng/mL in 84% of subjects",
                "AUC(0-12) > 1,807 h*ng/mL in 95% of subjects"),
  Threshold = c(3905, 2618, 1807),
  `Required percentile` = c("50th (median)", "16th", "5th"),
  Simulated = c(stats::quantile(auc_all, 0.50),
                stats::quantile(auc_all, 0.16),
                stats::quantile(auc_all, 0.05))
) |>
  dplyr::mutate(`Criterion met` = ifelse(Simulated > Threshold, "yes", "no"))

knitr::kable(bridging, digits = 0,
             caption = "Kamal 2014 pharmacokinetic bridging criteria versus the simulated 3 mg/kg twice-daily cohort.")

# All three are asserted at the published thresholds. Unlike the per-subgroup
# medians above, these are whole-cohort quantiles over 200 subjects and they
# clear their thresholds with margin: across 300 independent eta draws the
# worst-case p50, p16 and p05 sat 10%, 13% and 26% above 3,905, 2,618 and
# 1,807 respectively, and all three criteria were met in every draw.
stopifnot(
  stats::quantile(auc_all, 0.50) > 3905,
  stats::quantile(auc_all, 0.16) > 2618,
  stats::quantile(auc_all, 0.05) > 1807
)
```

All three criteria are met, reproducing the paper's conclusion that a
3 mg/kg twice-daily regimen is appropriate across the whole 0-12 month range.
Applying the same criterion to the lower doses the paper also simulated
shows why they were rejected - 2 mg/kg misses the 50% criterion outright and
2.5 mg/kg clears it only marginally, which is the quantitative form of the
paper's statement that "simulation of the lower doses of 2.5 and 2 mg/kg
showed a tendency toward lower exposures than benchmark comparators":

```{r bridging-lower-doses}
lower_doses <- tibble::tibble(
  `Dose (mg/kg BID)` = dose_levels,
  `Median AUC(0-12) (h*ng/mL)` = stats::median(auc_all) * dose_levels / 3
) |>
  dplyr::mutate(
    `Margin over 3,905 criterion` = sprintf(
      "%+.0f%%", 100 * (`Median AUC(0-12) (h*ng/mL)` / 3905 - 1)),
    `Exceeds 3,905 criterion` =
      ifelse(`Median AUC(0-12) (h*ng/mL)` > 3905, "yes", "no")
  )

knitr::kable(lower_doses, digits = c(1, 0, 0, 0),
             caption = "Cohort median steady-state OC AUC(0-12) across the simulated dose levels (exact dose-proportional scaling).")

stopifnot(
  # 2 mg/kg fails the median criterion; 3 and 3.5 mg/kg clear it. 2.5 mg/kg
  # sits just above and is deliberately not asserted either way -- it is
  # within the noise of the cohort draw.
  lower_doses$`Exceeds 3,905 criterion`[lower_doses$`Dose (mg/kg BID)` == 2]   == "no",
  lower_doses$`Exceeds 3,905 criterion`[lower_doses$`Dose (mg/kg BID)` == 3]   == "yes",
  lower_doses$`Exceeds 3,905 criterion`[lower_doses$`Dose (mg/kg BID)` == 3.5] == "yes"
)
```

The side-by-side comparison of the simulated cohort median against the
bridging comparator is below. The comparator is the 1-2 year-old reference
exposure, not a same-population NCA value, so a positive difference is the
expected result rather than a discrepancy: Kamal 2014 explicitly reports
"a tendency for predicted OC exposures in infants following a 3 mg/kg dose
to exceed those observed in adults receiving 75 mg twice daily and children
aged 1-5 years receiving approved dosages".

```{r nca-compare}
simulated_wide <- tibble::tibble(
  cohort  = "OC, infants, 3 mg/kg BID, steady state",
  auclast = stats::median(auc_all) / 1000       # h*ng/mL -> mg*h/L
)
reference_wide <- tibble::tibble(
  cohort  = "OC, infants, 3 mg/kg BID, steady state",
  auclast = 3905 / 1000
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = simulated_wide,
  reference = reference_wide,
  by        = "cohort",
  units     = c(auclast = "mg*h/L"),
  tolerance_pct = 20
)
knitr::kable(cmp, digits = 3,
             caption = paste("Simulated cohort median OC AUC(0-12) at steady state versus",
                             "the Kamal 2014 bridging comparator (mean AUCinf of 3,905 h*ng/mL",
                             "in children aged 1-2 years given 30 mg). * marks a difference",
                             "greater than 20%; here the exceedance is the paper's own finding."))
attr(cmp, "footnote")
```

For orientation against the other exposures the paper cites: twice-daily
75 and 150 mg oseltamivir in adults produced OC AUCs of about
3,000-6,000 h*ng/mL; 2 mg/kg twice daily gave a mean steady-state OC AUC of
2,969 h*ng/mL in children aged 1-12 years and 4,534 h*ng/mL in adolescents
aged 13-18 years. The simulated infant cohort median of
`r format(round(stats::median(auc_all)), big.mark = ",")` h*ng/mL sits inside
the adult 75-150 mg twice-daily band, which is the bridging conclusion the
paper draws.

# Assumptions and deviations

- **Virtual cohort demographics are synthetic.** Subject-level data are not
  public. Postnatal ages are sampled uniformly within the five Kamal 2014
  Table 1 enrolment bands using the published band proportions, and body
  weight is drawn from `WT = (3.2 + 0.14 * age_weeks) * exp(N(0, 0.13^2))`
  truncated to the observed 2.9-12.4 kg range. The intercept, slope and
  residual CV were chosen so that the simulated marginal weight and age
  distributions reproduce the Table 2 statistics (mean 6.5 kg / SD 2.1 and
  mean 23.5 weeks / SD 14.9); the reproduction is asserted in the cohort
  chunk. This is a calibrated approximation to a joint age-weight
  distribution the paper reports only marginally, not a resampling of the
  real cohort.
- **Age covariate is mapped to canonical `PNA` (months), not `AGE`
  (years).** Kamal 2014's covariate list explicitly contrasts "age" with
  "postconceptual age", which is exactly the register's `PNA` (postnatal)
  versus `PAGE` (postmenstrual) distinction, and the cohort spans only
  1.9-49.9 weeks, where a years scale loses resolution. The source column is
  in weeks with a 24-week reference; the model reparameterises the reference
  in place as `24 * 7 / 30.4375` months so the published constant stays
  visible. This is the same days-to-months reparameterisation the covariate
  register documents for `Zhao_2018_omeprazole`.
- **A single coefficient drives both OC parameters.** The Table 3 footnote
  states `VM,AGE = CLM,AGE`, so the model uses one parameter
  (`e_age_cl_oselcarb`) in both the `CLM/F` and `VM/F` equations rather than
  two numerically identical ones. Re-fitting this model to new data will
  therefore preserve the published constraint; a user who wants to relax it
  must split the parameter explicitly.
- **Allometric exponents are `fixed()`, not estimated.** Kamal 2014 Methods
  states the powers were "fixed power coefficients of 0.75 and 1" and that
  "Body weight dependencies were therefore fixed mechanistically" because
  weight and age were too strongly correlated to estimate simultaneously.
  Neither exponent appears among theta1-theta11 in Table 3, which is the
  confirming signal.
- **Reference weight 8 kg is a re-centering constant, not the cohort mean.**
  The Table 3 footnote and the Figure 2 caption both give 8 kg (the caption
  calls it "the median body weight"), while Table 2 reports a cohort mean of
  6.5 kg (SD 2.1). The two statements are not reconciled in the publication.
  The 8 kg value is used because it is the constant that appears in the
  published equations, and it is the only choice consistent with the
  tabulated `CL/F` and `CLM/F` estimates. As the Discussion notes, "Scaling
  to an adult weight (70 kg) instead of the median weight would not have
  altered the model because the power of the weight dependence was fixed",
  so the choice re-centres the typical values but does not change any
  prediction once weight is supplied.
- **No additive residual term for the parent.** The Methods section says
  "combined proportional and additive error models were used for oseltamivir
  and OC", but that describes the base model. Table 3 enumerates theta1
  through theta11 and contains no `sigmaOP,ADD` row, so the final model has
  proportional-only residual error for oseltamivir. The final estimates table
  is taken as authoritative and only `propSd` is encoded for the parent.
- **Figure 2's caption writes the normalisation as a multiplication.** The
  caption reads "OC clearance normalized by allometrically scaled body weight
  was derived using CLM*(WT/8)^0.75". Normalising by a factor is division,
  and only division leaves the linear age relationship the figure is drawn to
  demonstrate ("confirms the effect of maturation estimated by the model");
  multiplying would leave a residual weight dependence that dominates the
  plot. The replication above therefore divides. This affects only the figure
  reproduction - no model parameter depends on it.
- **The paper's claim of greater variability in the youngest subgroup is not
  reproduced, by construction.** Kamal 2014 notes "greater interindividual
  variability was seen in the youngest subjects (0-1 month) than in the other
  groups". The packaged model carries a single homogeneous 35.7% CV random
  effect on `CLM/F` at every age, so a synthetic cohort cannot generate an
  age-dependent spread. The published observation reflects the actual
  cohort's demographics - only 13 subjects were aged 30 days or less - rather
  than a structural feature of the model.
- **The age effect is asserted on typical values, not on stochastic subgroup
  medians.** Adjacent age subgroups differ by only 6-10% in typical OC
  exposure, while the random effect on `CLM/F` has a 35.7% CV; with 20-53
  simulated subjects per subgroup the sample medians are therefore strictly
  monotone in only about 40% of random draws, even though the underlying
  model is exactly monotone. The quantitative Figure 4 gate is consequently
  placed on the zero-random-effect predictions (which depend only on the
  base-R-seeded covariates and are exactly reproducible), and the stochastic
  cohort is asserted for direction alone. The whole-cohort bridging quantiles
  in the next section are a different matter: they pool all 200 subjects and
  clear their thresholds in every one of 300 test draws, so they are asserted
  at the published values.
- **Complete conversion means OC exposure is mass-balanced against the parent
  dose.** With fm fixed at 1, steady-state OC AUC equals `Dose/(CLM/F)`
  exactly. No molecular-weight correction is applied between oseltamivir
  (312.4 g/mol) and OC (284.4 g/mol); the ratio is absorbed into the
  "apparent" `CLM/F` and `VM/F` scaling, matching the original NONMEM
  implementation.
- **Steady state is not fully attained in five days.** The parent's terminal
  half-life is about 15 h, so the five-day (ten-dose) regimen the paper
  simulated reaches roughly 99.4% of steady state. The structural closed-form
  gates therefore use a ten-day regimen, while the exposure and bridging
  results use the paper's five-day regimen and are asserted against the
  closed form with a 1% tolerance.
- **Do not extrapolate beyond one year of age.** Kamal 2014 Discussion states
  that the age dependence "should not be extrapolated beyond the range of the
  analysis data (<1 year)", and that generalisability to extremely premature
  neonates "may be suboptimal" because infants of postmenstrual age below
  36 weeks were excluded from both trials. The linear age term has no upper
  plateau, so extrapolation past 12 months will inflate `CLM/F` and `VM/F`
  without bound.
- **Screened-but-rejected covariates carry no point estimates.**
  Postconceptual age, sex, and a CASG114 relative-bioavailability
  (formulation) effect were all tested in the full covariate model and none
  was retained. They are documented in the model file's
  `covariatesDataExcluded` list for provenance; no coefficient is available
  for any of them, and relative bioavailability is 1 for both studies.
- **Table 2's ethnicity rows are internally inconsistent in the source.**
  As printed, Hispanic 34 (25.6%), Non-Hispanic/Latino 93 (69.9%), Other 12
  (9.0%) and Unknown 6 (4.5%) sum to 145 subjects and 109%, exceeding the
  analysis N of 133. The values are recorded verbatim in the model's
  `population$notes`; the race rows are consistent (105 + 14 + 12 + 2 = 133)
  and are the ones carried into `population$race_ethnicity`. Neither race nor
  ethnicity was screened as a covariate, so nothing in the model depends on
  the discrepancy.
