---
title: "Meltdose tacrolimus (Kamp 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Meltdose tacrolimus (Kamp 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Kamp_2023_tacrolimus"))$reference`
* Description: `r rxode2::rxode(readModelDb("Kamp_2023_tacrolimus"))$description`
* Article: <https://doi.org/10.3390/pharmaceutics16010017>
* Supplement (figures S1-S6, tables S1-S3 and the NONMEM control stream,
  Supplementary Data S1):
  <https://www.mdpi.com/article/10.3390/pharmaceutics16010017/s1>

Meltdose tacrolimus (Envarsus) is a prolonged-release oral formulation. Kamp
2023 developed a population PK model for it in elderly de novo kidney
transplant recipients, then used the model to derive limited sampling
strategies for AUC estimation and to quantify how long meltdose tacrolimus
takes to reach steady state.

# Population

Thirty-four de novo kidney transplant recipients aged 65 years or older were
analysed (36 enrolled, 2 excluded after their transplantation was postponed).
Median age was 71.5 years (IQR 68.8-73.2) and median weight 77.0 kg (IQR
68.1-84.5); 12 (35.3%) were female. All patients started on meltdose
tacrolimus 7 mg once daily on the day of transplantation, with subsequent
AUC-guided dose individualisation; the median dose at AUC sampling was 6.0 mg
(IQR 3.25-7.0). CYP3A5 genotype was *3/*3 in 26 (76.5%), *1/*3 in 6 (17.6%)
and *1/*1 in 2 (5.9%). The dataset comprised 546 concentrations over 87 AUC
profiles (median 2 per subject, range 1-5); 37 of those profiles (42.5%) were
abbreviated dried blood spot profiles collected at home by 20 (58.8%) of the
patients. Baseline characteristics are Kamp 2023 Table 1.

The same information is available programmatically from the model's
`population` metadata:

```{r population}
str(rxode2::rxode(readModelDb("Kamp_2023_tacrolimus"))$population)
```

# Source trace

Per-parameter provenance is recorded as an in-file comment beside each `ini()`
entry in `inst/modeldb/specificDrugs/Kamp_2023_tacrolimus.R`. Collected here
for review. "S1" is Supplementary Data S1, the paper's NONMEM control stream.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lktr` (Ktr) | 0.752 1/h | Table 2 "Ktr (h-1)", RSE 21%; S1 `$THETA(6)` |
| `ltlag` (lag, lagged class) | 2.29 h | Table 2 "Lag (h)", RSE 2%; S1 `$THETA(5)` |
| `lcl` (CL/F at 70 kg) | 19.6 L/h | Table 2 "CL/F", RSE 8%; S1 `$THETA(7)` |
| `lvc` (Vc/F at 70 kg) | 123 L | Table 2 "Vc/F", RSE 38%; S1 `$THETA(8)` |
| `lq` (Q/F at 70 kg) | 74.9 L/h | Table 2 "Q/F", RSE 9%; S1 `$THETA(9)` |
| `lvp` (Vp/F at 70 kg) | 500 L, fixed | Table 2 "Vp/F (Fixed)"; Sect. 3.3.1 (not identifiable; fixed to the literature values of Kamp refs [10] Martial 2021 and [20] Benkali 2010); S1 `$THETA(10) 500 FIX` |
| `lfdepot` (F) | 1, fixed | Table 2 "F (Fixed)"; Sect. 2.3.1 |
| `e_wt_cl_q` | 0.75, fixed | Sect. 2.3.2 allometric equation; S1 `ALLO_CL = ALLO_V ** 0.75` |
| `e_wt_vc_vp` | 1, fixed | Sect. 2.3.2 allometric equation; S1 `ALLO_V = WT/70` |
| mixture proportion (no-lag class) | 26% | Table 2 "Pop parameter (%)", RSE 50%; S1 `$THETA(11)` (see Assumptions) |
| `etalktr` | 0.312 | S1 `$OMEGA`; Table 2 IIV Ka 60.5% CV = sqrt(exp(0.312) - 1) |
| `etalcl` | 0.0979 | S1 `$OMEGA BLOCK(2)`; Table 2 IIV CL 32.1% CV |
| `cov(etalcl, etalvc)` | 0.23 | S1 `$OMEGA BLOCK(2)`; correlation 0.23/sqrt(0.0979*0.609) = 0.94, quoted as "94% covariance" in Sect. 3.3.1 |
| `etalvc` | 0.609 | S1 `$OMEGA BLOCK(2)`; Table 2 IIV Vc 91.6% CV |
| `etaiov_cl_1` .. `etaiov_cl_5` | 0.224 | S1 `$OMEGA BLOCK(1)` + 4 x `SAME`; Table 2 IOV CL 50.1% CV |
| `propSdVenous` | 0.208 | Table 2 "Proportional Error Whole blood", RSE 8%; S1 `$THETA(1)` |
| `propSdCapillary` | 0.307 | Table 2 "DBS", RSE 12%; S1 `$THETA(2)` |
| `addSd` | 1e-4 ug/L, fixed | S1 `$THETA(3) 0.0001 FIX` (numerical stabiliser; not in Table 2) |
| ODE chain `depot -> transit1 -> transit2 -> central` at `ktr` | n/a | S1 `$MODEL` / `$DES`; Figure 2 |
| `central <-> peripheral1` via `k12`/`k21` | n/a | S1 `$PK` `K45 = Q/V2`, `K54 = Q/V3` |
| `alag(depot) <- tlag * MIX_LAGGED_ABS` | n/a | S1 `$MIX` / `$PK` `ALAG1`; Sect. 3.3.1 |
| `Cc <- 1000 * central / vc` | n/a | S1 `$PK` `S4 = V2/1000` (mg and L in, ug/L out) |
| per-record residual switch | n/a | S1 `$ERROR` `IF(DBS0.EQ.1)`; Sect. 3.3.1 |

The final model retained **no** clinical covariate. Hematocrit and CYP3A5
genotype were selected by the stepwise search but rejected as clinically
irrelevant (Sect. 3.3.2); they are documented in the model file's
`covariatesDataExcluded` metadata together with age, sex, prednisolone dose,
calcium-channel-blocker use and diabetes.

# Virtual cohort

Original observed data are not publicly available. The cohort below mirrors
the simulation Kamp 2023 itself ran for its time-to-steady-state analysis
(Sect. 2.3.4): a 70 kg individual receiving 7 mg meltdose tacrolimus once
daily, followed for 14 days. The paper simulated 1000 profiles; 200 is used
here (the per-arm cap for these vignettes) and is ample for the median and
quartile summaries the paper reports.

```{r cohort}
# `set.seed()` seeds R's RNG (used for the Bernoulli mixture draw below). It
# does NOT seed rxode2's simulation RNG, whose streams are partitioned per
# solver thread -- so this cohort differs between a 2-core CI runner and a
# 16-thread workstation. Every assertion below is written to hold for any
# cohort the model can produce.
set.seed(2024)

n_sub  <- 200L
tau    <- 24      # dosing interval (h)
n_dose <- 14L     # 14 daily doses -> the paper's day-14 steady-state anchor
t_end  <- tau * n_dose

subjects <- tibble(
  id = seq_len(n_sub),
  WT = 70,
  OCC = 1L,
  # The lagged class is the 74% majority; see "Assumptions and deviations".
  MIX_LAGGED_ABS = rbinom(n_sub, 1L, 0.74),
  SAMPLE_CAPILLARY = 0L
)

# Hourly over the whole course, refined to 15 min over the first and last
# dosing intervals, which are the two windows any Cmax / Tmax is read from.
obs_times <- sort(unique(c(
  seq(0, t_end, by = 1),
  seq(0, tau, by = 0.25),
  seq(t_end - tau, t_end, by = 0.25)
)))

events <- rxode2::et(amt = 7, ii = tau, addl = n_dose - 1L, cmt = "depot") |>
  rxode2::et(obs_times) |>
  rxode2::et(id = seq_len(n_sub)) |>
  as.data.frame() |>
  dplyr::left_join(subjects, by = "id")

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))

# Realised share of the lagged absorption class (target 74%).
sprintf("%.0f%% of the cohort drew the lagged-absorption class",
        100 * mean(subjects$MIX_LAGGED_ABS))
```

# Simulation

```{r simulate}
mod <- readModelDb("Kamp_2023_tacrolimus")
sim <- rxode2::rxSolve(
  mod, events = events,
  maxsteps = 200000L,          # 14 lagged daily doses exhaust the default
  addDosing = FALSE,
  keep = c("MIX_LAGGED_ABS")
) |>
  as.data.frame()
```

# Replicate Figure 4C: median concentration-time profile

```{r figure-4c}
# Replicates Figure 4C of Kamp 2023: median simulated tacrolimus
# concentration (IQR shaded) vs. time after tacrolimus initiation, 70 kg
# individual on 7 mg meltdose tacrolimus q.d.
prof <- sim |>
  group_by(time) |>
  summarise(
    Q25 = quantile(Cc, 0.25),
    Q50 = median(Cc),
    Q75 = quantile(Cc, 0.75),
    .groups = "drop"
  )

ggplot(prof, aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q25, ymax = Q75), fill = "grey70", alpha = 0.5) +
  geom_line(linewidth = 0.4) +
  scale_x_continuous(breaks = seq(0, t_end, by = 48)) +
  labs(x = "Time after tacrolimus initiation (h)",
       y = "Tacrolimus concentration (ug/L)",
       caption = "Replicates Figure 4C of Kamp 2023.")
```

The published panel rises to a first-dose median peak just above 11 ug/L,
accumulates over roughly the first week, and settles into a steady-state
interval whose nadir sits a few hours *after* each dose -- the visible
signature of the absorption lag -- before rising to a peak near 20 ug/L.

```{r figure-4c-checks}
day1 <- prof |> filter(time <= tau)
ss   <- prof |> filter(time >= t_end - tau, time <= t_end)

fig4c <- tibble(
  quantity = c("Day-1 median Tmax (h)", "Day-1 median Cmax (ug/L)",
               "Steady-state median peak (ug/L)",
               "Steady-state median nadir (ug/L)"),
  simulated = c(day1$time[which.max(day1$Q50)], max(day1$Q50),
                max(ss$Q50), min(ss$Q50)),
  # On-screen digitisation of the published panel (see Assumptions). The
  # x-axis resolves ~2 pixels per hour, so times carry about +/- 0.5 h and
  # concentrations a few percent of systematic y-calibration error.
  figure_4c = c(6.08, 11.27, 20.55, 10.27)
)
fig4c |>
  dplyr::rename("Quantity" = quantity,
                "Simulated" = simulated,
                "Figure 4C (digitised)" = figure_4c) |>
  knitr::kable(digits = 2,
               caption = paste("Figure 4C reference values are read off the",
                               "published panel by on-screen digitisation and",
                               "are approximate (see Assumptions)."))

# The three concentrations landed +11 / +10 / +8% against the digitised panel
# in one cohort and -4 / -7 / -7% in another, so the gap is cohort redraw
# rather than a systematic offset in either direction. 30% is roughly 3x the
# largest deviation seen and still goes red on a mis-transcribed clearance,
# dose or scale factor, each of which moves these by tens of percent.
stopifnot(max(abs(fig4c$simulated[2:4] / fig4c$figure_4c[2:4] - 1)) < 0.30)

# Tmax is the one quantity that discriminates the two readings of the
# absorption-lag mixture (see Assumptions): 6.0 h under the adopted
# 74%-lagged reading vs 4.25 h under the alternative. 1.5 h keeps the gate
# able to detect that flip while admitting the cohort-to-cohort wobble of a
# 200-subject median read on a 15-minute grid.
stopifnot(abs(fig4c$simulated[1] - fig4c$figure_4c[1]) < 1.5)
```

# Replicate Figure 4A/B: time to steady state

Kamp 2023 Sect. 2.3.4 defines the fraction of steady state as
`FSS = IPRED / Ctrough_14 * 100`, where `Ctrough_14` is the trough
concentration on day 14, and evaluates it at the daily troughs. The reported
results (Sect. 3.3.4) are a median time to 90% of steady state of 96 h (IQR
72-120) and to 97% of 144 h (IQR 120-312), with 76% / 40% of subjects at 90% /
97% of steady state five days after initiation and 94% / 71% after seven days.

```{r figure-4ab}
troughs <- sim |>
  filter(time %in% seq(tau, t_end, by = tau)) |>
  select(id, time, Cc) |>
  group_by(id) |>
  mutate(fss = 100 * Cc / Cc[time == t_end]) |>
  ungroup()

first_time_at <- function(threshold) {
  troughs |>
    filter(fss >= threshold) |>
    group_by(id) |>
    summarise(t = min(time), .groups = "drop")
}
t90 <- first_time_at(90)
t97 <- first_time_at(97)

prob_at <- function(threshold, when) {
  100 * mean(troughs$fss[troughs$time == when] >= threshold)
}

ss_tab <- tibble(
  claim = c("Median time to 90% of steady state (h)",
            "Median time to 97% of steady state (h)",
            "P(>= 90% of steady state) at day 5 (%)",
            "P(>= 97% of steady state) at day 5 (%)",
            "P(>= 90% of steady state) at day 7 (%)",
            "P(>= 97% of steady state) at day 7 (%)"),
  published = c(96, 144, 76, 40, 94, 71),
  simulated = c(median(t90$t), median(t97$t),
                prob_at(90, 120), prob_at(97, 120),
                prob_at(90, 168), prob_at(97, 168))
)
ss_tab |>
  dplyr::rename("Claim" = claim,
                "Published (Kamp 2023)" = published,
                "Simulated" = simulated) |>
  knitr::kable(digits = 1,
               caption = "Time to steady state: Kamp 2023 Sect. 3.3.4 vs. simulation.")
```

```{r figure-4ab-gate}
# The two medians are read off a 24-h grid, so they can only land on a
# multiple of 24; one grid step either side is the tightest bound that can
# hold across cohorts.
stopifnot(
  abs(ss_tab$simulated[1] - 96)  <= 24,
  abs(ss_tab$simulated[2] - 144) <= 48
)
# The four probabilities are cohort proportions, and they move noticeably
# between cohorts: the largest gap to a published value was 2 points in one
# 200-subject draw and 8 points in another. The binomial standard error at
# n = 200 near p = 0.4 is about 3.5 points, and Kamp 2023 simulated 1000
# subjects, so some of that gap is irreducible. 12 points sits outside the
# observed range with headroom and still goes red on a structural error (a
# wrong CL or Vp moves these by 30 points or more). Do not tighten this to
# whatever a single local run happens to give.
stopifnot(max(abs(ss_tab$simulated[3:6] - ss_tab$published[3:6])) < 12)
```

```{r figure-4b-plot}
# Replicates Figure 4B of Kamp 2023: probability of having reached at least
# 90% (blue) and 97% (black) of the steady-state concentration.
troughs |>
  group_by(time) |>
  summarise(`90%` = 100 * mean(fss >= 90),
            `97%` = 100 * mean(fss >= 97), .groups = "drop") |>
  pivot_longer(-time, names_to = "threshold", values_to = "probability") |>
  ggplot(aes(time, probability, colour = threshold)) +
  geom_step(linewidth = 0.6) +
  scale_x_continuous(breaks = seq(0, t_end, by = 48)) +
  scale_colour_manual(values = c(`90%` = "#2C7FB8", `97%` = "black")) +
  labs(x = "Time after tacrolimus initiation (h)",
       y = "Steady state probability (%)", colour = "Fraction of\nsteady state",
       caption = "Replicates Figure 4B of Kamp 2023.")
```

# PKNCA validation

Kamp 2023 has no published NCA table. It does, however, define a reference
exposure used throughout the paper (Sect. 2.3.1 and 2.3.3, and the captions of
Figures S3 and S4):

`AUC0-24 = (F x D x 1000) / CL`, with F fixed at 1, D the dose in mg and CL the
apparent elimination clearance in L/h.

That identity is the yardstick every limited sampling strategy in the paper was
scored against, so reproducing it from a trapezoidal NCA of the simulated
steady-state interval is a direct check of the absorption chain's mass balance,
the bioavailability anchor and the `S4 = Vc/1000` concentration scaling. It is
evaluated at typical values (random effects zeroed) so the published CL of
19.6 L/h at 70 kg applies exactly, across the cohort's median dose and its
interquartile dose range (Table 1).

```{r pknca}
doses <- c(3.25, 6.0, 7.0)

make_dose_arm <- function(i) {
  rxode2::et(amt = doses[i], ii = tau, addl = n_dose - 1L, cmt = "depot") |>
    rxode2::et(seq(0, t_end, by = 0.25)) |>
    as.data.frame() |>
    mutate(id = i, WT = 70, OCC = 1L, MIX_LAGGED_ABS = 1L,
           SAMPLE_CAPILLARY = 0L,
           treatment = sprintf("%.2f mg q.d.", doses[i]))
}
events_tv <- bind_rows(lapply(seq_along(doses), make_dose_arm))
stopifnot(!anyDuplicated(unique(events_tv[, c("id", "time", "evid")])))

sim_tv <- rxode2::rxSolve(
  rxode2::zeroRe(rxode2::rxode(mod)), events = events_tv,
  maxsteps = 200000L, addDosing = FALSE, keep = c("treatment")
) |>
  as.data.frame()

# Restrict to the final dosing interval; the interval start is a real
# measurement, so no time-zero back-fill is needed here.
sim_nca <- sim_tv |>
  dplyr::filter(!is.na(Cc), time >= t_end - tau) |>
  dplyr::select(id, time, Cc, treatment)
stopifnot(nrow(sim_nca) > 0)

dose_df <- events_tv |>
  dplyr::filter(evid == 1) |>
  dplyr::group_by(id, treatment) |>
  dplyr::summarise(time = max(time), amt = dplyr::first(amt), .groups = "drop") |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "ug/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "mg")

intervals <- data.frame(
  start   = t_end - tau,
  end     = t_end,
  cmax    = TRUE,
  tmax    = TRUE,
  cmin    = TRUE,
  auclast = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
```

## Comparison against the paper's reference AUC

```{r nca_compare}
published <- tibble::tibble(
  treatment  = sprintf("%.2f mg q.d.", doses),
  # Kamp 2023 AUCref = (F * D * 1000) / CL with F = 1 and CL = 19.6 L/h at 70 kg.
  auclast    = doses * 1000 / 19.6
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "treatment",
  units         = c(auclast = "ug*h/L"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = paste(
  "Simulated steady-state AUC0-24 vs. the paper's reference AUC identity",
  "(F x D x 1000) / CL. * differs from reference by >20%."
))
```

```{r nca_gate}
auc_sim <- as.data.frame(nca_res$result) |>
  filter(PPTESTCD == "auclast") |>
  left_join(published, by = "treatment", suffix = c("", "_ref"))
pct <- 100 * (auc_sim$PPORRES / auc_sim$auclast - 1)

# Deterministic (typical-value) comparison: the only error left is the
# trapezoidal discretisation of a 0.25 h grid, so a tight bound is correct
# here and is what makes this gate useful. Realised 0.014% at every dose.
stopifnot(max(abs(pct)) < 0.5)

# Dose proportionality: AUCref is linear in dose, and so is the model.
ratio <- auc_sim$PPORRES / doses
stopifnot(max(ratio) / min(ratio) - 1 < 0.01)
```

Steady-state `Tmax` and `Cmin` from the same NCA give the other two readable
features of Figure 4C:

```{r nca_extra}
# PKNCA already reports tmax relative to the start of the interval, so it
# reads directly as hours after the 14th dose.
as.data.frame(nca_res$result) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "cmin")) |>
  select(treatment, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  rename("Dose group" = treatment,
         "Cmax,ss (ug/L)" = cmax,
         "Tmax,ss (h post dose)" = tmax,
         "Cmin,ss (ug/L)" = cmin) |>
  knitr::kable(digits = 2,
               caption = "Typical-value steady-state NCA over the 14th dosing interval.")
```

# Assumptions and deviations

* **Direction of the absorption-lag mixture (the one substantive judgement in
  this extraction).** Kamp 2023's source is self-contradictory about which
  class the estimated mixture proportion of 26% belongs to.

  - Table 2 and Table S2 footnote the parameter as "Pop parameter, proportion
    of the population belonging to the group **without** oral lag time", and
    the `$THETA` comments in Supplementary Data S1 agree with that reading
    (`0 FIX ; Absorption lag time for group 1`, with `P(1) = THETA(11) =
    0.26`). Under this reading 74% of subjects carry the 2.29 h lag.
  - The `$PK` block printed in the same supplement reads
    `IF(MIXNUM.EQ.2) ALAG1 = THETA(4) ELSE ALAG1 = THETA(5)`, which assigns
    the 2.29 h lag to `MIXNUM = 1` and therefore to the 26% probability --
    the opposite assignment, and inconsistent with its own `$THETA` comments.

  The **footnote reading (74% lagged)** is adopted, on the evidence of the
  paper's own Figure 4C. That panel is the median of 1000 profiles simulated
  from this final model for a 70 kg subject on 7 mg q.d., so it is a direct
  readout of the fitted mixture. Its median sits on the baseline for the first
  three hours and peaks at about 6.1 h (on-screen digitisation of the
  published panel; the x-axis resolution is roughly 2 pixels per hour, so
  treat this as +/- 0.5 h). Simulating both readings gives a day-1 median
  Tmax of 6.0-6.25 h under the 74%-lagged reading, versus 4.25 h under the
  26%-lagged reading -- the latter also puts the median at 5.4 ug/L by 2 h,
  where the published panel is still flat on zero. A majority-lagged
  population is additionally what makes the median profile show a lag at all:
  with only 26% lagged, the median subject has none, and the published panel
  would not show the flat first three hours or the post-dose nadir that it
  does.

  The consequence of getting this wrong is a roughly 2 h shift in Tmax and no
  change to AUC, so the steady-state and exposure checks above cannot
  discriminate the two readings -- only the shape checks can. Both the model
  file's `covariateData[[MIX_LAGGED_ABS]]$notes` and the
  `MIX_LAGGED_ABS` entry in `inst/references/covariate-columns.md` record the
  discrepancy so a future reader is not surprised by the supplement.

* **The mixture proportion is metadata, not a parameter.** Following the
  convention set by `Bonate_2004_apomine` and `Frymoyer_2013_mycophenolic_acid`,
  class membership enters as the binary `MIX_LAGGED_ABS` covariate and the 26%
  / 74% split is documented rather than estimated. To simulate a population,
  draw `MIX_LAGGED_ABS ~ Bernoulli(0.74)` per subject as the cohort chunk
  above does; for a typical-value profile, set it to 1.

* **Figure 4C reference values are digitised, not published.** The four
  numbers in the Figure 4C comparison table (Tmax, Cmax, steady-state peak and
  trough) were read off the published panel by on-screen digitisation; Kamp
  2023 prints none of them. They are used only as loose shape checks, with
  bounds set well outside the digitisation error. Every gate that uses a
  printed value -- the six time-to-steady-state claims and the reference AUC
  identity -- is tighter.

* **Occasion is held at 1.** The model carries five inter-occasion variability
  slots on CL because the paper's dataset had up to five AUC profiles per
  patient. Forward simulations of a single continuous treatment course have no
  natural occasion boundary, so `OCC = 1` is used throughout and the IOV
  contributes one draw per subject rather than switching mid-course. Set
  `OCC` to 1-5 per profile to reproduce the paper's within-subject
  occasion-to-occasion variability; values outside 1-5 switch all IOV
  indicators off and leave only the IIV.

* **The limited sampling strategies are not reproduced.** Table S3 and the
  paper's LSS conclusions (best three-sample scheme T = 0, 2, 6 h; best
  four-sample schemes T = 0, 2, 6, 8 h and T = 0, 1, 6, 8 h) come from
  maximum-a-posteriori Bayesian re-estimation of each simulated subject from a
  reduced sampling schedule. That is a fitting exercise rather than a
  simulation, and is outside what a packaged model plus a validation vignette
  can check. The structural model that the LSS analysis rests on is what is
  validated here.

* **The alternative covariate model is not encoded.** Table S2 reports a
  second parameter set including hematocrit (-4.94) and CYP3A5 expresser
  (0.625) effects on CL. Kamp 2023 rejected both as clinically irrelevant
  (Sect. 3.3.2) and reports the covariate-free model as final, so the final
  model is what is packaged. The screened covariates are recorded in the model
  file's `covariatesDataExcluded` metadata.

* **Whole blood is the reference matrix.** `SAMPLE_CAPILLARY = 0` is used
  throughout, which selects the 20.8% whole-blood proportional residual error.
  Set it to 1 on records collected as dried blood spots to select the 30.7%
  term. The choice affects only the residual error, never the structural
  prediction: Kamp 2023 modelled DBS and whole blood as measurements of the
  same underlying concentration differing in precision. The switch is written
  arithmetically inside `model()`
  (`propSd <- propSdCapillary * SAMPLE_CAPILLARY + propSdVenous * (1 -
  SAMPLE_CAPILLARY)`), following the `SAMPLE_CAPILLARY` register entry and its
  founding example `Cleary_2023_risdiplam`.

* **The additive residual term is a stabiliser.** `addSd` is fixed at
  1e-4 ug/L, matching `$THETA(3) 0.0001 FIX` in the control stream, where it
  exists only to keep the residual weight away from zero. It is not reported
  in Table 2 and is numerically irrelevant on a ug/L scale.

* **Fitting-dataset profiles were not all at steady state.** The first AUC was
  obtained a median of 7 days after transplantation, and the paper's own
  analysis concludes that 5-7 days are needed to reach 90% of steady state.
  The model was fitted to the full time course rather than to steady-state
  profiles, so it is appropriate for de novo dosing simulations from day 0 --
  which is how it is used above.
