---
title: "Quizartinib and AC886 (Kang 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quizartinib and AC886 (Kang 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-load}
mod <- readModelDb("Kang_2020_quizartinib")
ui <- rxode2::rxode(mod)
```

* Citation: `r ui$reference`
* Article: <https://doi.org/10.1002/jcph.1680> (open access, PMC7689835)
* Supplement: goodness-of-fit Figures S1-S4 only; all parameter values come
  from the main article (Tables 3 and 4, equations 1-9, Figure 2).

`r ui$description`

A later analysis of the same drug in newly diagnosed AML (QuANTUM-First) by the
same sponsor is available as `Vaddady_2024_quizartinib`; it re-estimated the
model on a larger pooled dataset and uses body weight rather than body surface
area, so the two are separate models.

# Population

Kang 2020 pooled 649 participants from 8 studies (Table 1): 325 healthy
volunteers from five phase 1 single-dose studies (relative bioavailability and
dose proportionality AC220-014; ketoconazole/fluconazole interaction AC220-015;
hepatic impairment AC220-016; lansoprazole interaction AC220-018; food effect
AC220-019) and 324 patients with relapsed/refractory AML from a phase 1
maintenance study (2689-CL-0011), the phase 2b dose-ranging study
(2689-CL-2004) and the phase 3 QuANTUM-R study (239 patients). Healthy
volunteers received single doses of 30-90 mg; patients received once-daily
doses of 20-90 mg. Median age was 33 years in healthy volunteers and 55 years
in patients; median body surface area was 1.9 m^2 in both; 42.1% were female
and 18.0% were Black or African American (mostly among the healthy volunteers).
Median albumin was 4.4 g/dL in healthy volunteers and 3.7 g/dL in patients.
Strong CYP3A inhibitors were used by 8.9% of healthy volunteers (the
ketoconazole arm) and 28.7% of patients (Table 2). The analysis used 11,770
quizartinib and 10,888 AC886 concentrations.

```{r population}
str(ui$population)
```

# Source trace

| Element | Value | Source |
|---|---|---|
| Structure: 3-cmt quizartinib, zero-order input to depot over D1 then first-order ka, lag time | -- | Results "Quizartinib Population PK Model"; Figure 2; Table 3 row 'D1 (duration of zero-order input to depot compartment)' |
| Structure: 2-cmt AC886 fed by fMET x CL | -- | Figure 2; Results "AC886 Population PK Model" |
| `lcl` | 2.77 L/h | Table 3; eq 4 |
| `e_dis_aml_cl`, `e_cyp3a4_inh_strong_cl` | 0.0820, -0.441 (additive bracket) | Table 3; eq 4 |
| `lvc`, `e_alb_vc`, `e_bsa_vc` | 194 L; (ALB/4.1)^-0.725; (BSA/1.9)^1.46 | Table 3; eq 5 |
| `lq`, `e_bsa_q` | 27.9 L/h; (BSA/1.9)^0.970 | Table 3; eq 7 |
| `lvp`, `e_bsa_vp`, `e_age_vp` | 170 L; (BSA/1.9)^1.50; (AGE/44)^0.453 | Table 3; eq 6 |
| `lq2`, `lvp2` | 0.567 L/h, 39.3 L | Table 3 |
| `lka`, `lka_aml`, `e_fed_ka` | 0.874 1/h, 1.68 1/h, -0.512 | Table 3; eq 1 |
| `ld1`, `ltlag` | 0.708 h, 0.205 h | Table 3 |
| `lfdepot`, `e_dis_aml_fdepot`, `e_cyp3a4_inh_strong_fdepot` | 1 (fixed), 0.599, 0.136 | eq 2; Table 3 |
| `lfdepot_ac220019`, `e_fed_fdepot` | 0.913, 0.0509 | eq 3; Table 3 |
| `lcl_ac886`, `e_bsa_cl_ac886` | 4.09 L/h; (BSA/1.9)^1.60 | Table 4; eq 9 |
| `e_race_black_cl_ac886`, `e_cyp3a4_inh_strong_cl_ac886` | 0.586, 0.106 (additive bracket) | Table 4; eq 9 |
| `lvc_ac886`, `e_cyp3a4_inh_strong_vc_ac886` | 4.95 L; 1.92 | Table 4; eq 8 |
| `lvp_ac886`, `lq_ac886` | 70.6 L, 3.29 L/h | Table 4 |
| `fmet` | 0.5 (fixed) | Table 4; Results "Model Building" |
| IIV CL, Vc, Q1, Vp1, ka, D1, F1, ALAG1 | 55.1, 27.6, 24.8, 39.7, 38.5, 69.3, 34.8, 62.0% CV | Table 3 |
| IIV Vcm; CLm healthy / patients | 117% CV; 45.9% / 64.1% CV | Table 4 |
| IOV F1 (occasions 1, 3); CL, Vc (occasions 1-4) | 22.6, 40.9, 20.3% CV | Table 3 |
| Quizartinib residual: proportional variance healthy / patients; additive variance | 0.00563 / 0.0376; 0.956 | Table 3 and footnotes b, c |
| AC886 residual (log scale) healthy / patients | variance 0.0690 / 0.169 (SD 0.263 / 0.412) | Table 4 and footnote a |

# Helper functions

Two declared endpoints (`Cc` and `Cc_ac886`) place the endpoint slots after
the six ODE states, so observation rows carry `dvid = 1` with no `cmt`; both
concentrations are returned as columns on every row. Doses carry `rate = -2`
so the modelled zero-order duration D1 is applied.

```{r helpers}
make_events <- function(cohort, dose, n_doses, obs_times) {
  doses <- cohort |>
    dplyr::mutate(
      time = 0, amt = dose, evid = 1L, rate = -2, ii = 24,
      addl = as.integer(n_doses - 1), cmt = "depot", dvid = NA_integer_
    )
  obs <- tidyr::crossing(cohort, time = obs_times) |>
    dplyr::mutate(
      amt = NA_real_, evid = 0L, rate = NA_real_, ii = 0, addl = 0L,
      cmt = NA_character_, dvid = 1L
    )
  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    as.data.frame()
}

cov_cols <- c(
  "DIS_AML", "CONMED_CYP3A4_INH_STRONG", "ALB", "BSA", "AGE",
  "RACE_BLACK", "FED", "STUDY_AC220019", "OCC"
)

# Kang 2020 Figure 4 reference patient: AML, no strong CYP3A inhibitor,
# albumin 3.7 g/dL (37 g/L), BSA 1.9 m^2. Age does not affect AUC; the AML
# median age 55 years is used.
ref_patient <- data.frame(
  DIS_AML = 1, CONMED_CYP3A4_INH_STRONG = 0, ALB = 37, BSA = 1.9, AGE = 55,
  RACE_BLACK = 0, FED = 0, STUDY_AC220019 = 0, OCC = 1
)
```

# Covariate effects at steady state (replicates Figure 4)

Kang 2020 Figure 4 compares steady-state `AUC0-24,ss` and `Cmax,ss` for AML
patients on 60 mg once daily at the 5th/25th/75th/95th percentiles of each
continuous covariate, and for each categorical level, against the reference
patient. The figure shows medians of 1000 simulations with parameter
uncertainty; the typical-value ratios below correspond to those medians. The
paper does not state the simulation day. A 28-day solve (one treatment cycle)
is used here; the analytic steady state is checked separately in the next
section.

```{r forest-sim}
scen <- dplyr::bind_rows(
  ref_patient |> dplyr::mutate(scenario = "Reference"),
  ref_patient |> dplyr::mutate(scenario = "Strong CYP3A inhibitor", CONMED_CYP3A4_INH_STRONG = 1),
  ref_patient |> dplyr::mutate(scenario = "Black race", RACE_BLACK = 1),
  dplyr::bind_rows(lapply(c(2.7, 3.3, 4.0, 4.4), function(a) {
    ref_patient |> dplyr::mutate(scenario = sprintf("Albumin %.1f g/dL", a), ALB = a * 10)
  })),
  dplyr::bind_rows(lapply(c(1.5, 1.7, 2.0, 2.3), function(b) {
    ref_patient |> dplyr::mutate(scenario = sprintf("BSA %.1f m^2", b), BSA = b)
  })),
  dplyr::bind_rows(lapply(c(25, 45, 65, 74), function(a) {
    ref_patient |> dplyr::mutate(scenario = sprintf("Age %d y", a), AGE = a)
  }))
) |>
  dplyr::mutate(id = dplyr::row_number())

obs_forest <- sort(unique(c(seq(0, 24, by = 0.25), seq(27 * 24, 28 * 24, by = 0.1))))
ev_forest <- make_events(scen, dose = 60, n_doses = 28, obs_times = obs_forest)

sim_forest <- rxode2::rxSolve(
  rxode2::zeroRe(mod), ev_forest,
  keep = c("scenario"), returnType = "data.frame", useLinCmt = FALSE
)
```

The exposure metrics are computed with PKNCA over the last dosing interval
(648-672 h).

```{r forest-nca}
nca_forest <- function(sim, conc_col) {
  conc <- sim |>
    dplyr::filter(!is.na(.data[[conc_col]])) |>
    dplyr::transmute(scenario, id, time, conc = .data[[conc_col]])
  dose <- ev_forest |>
    dplyr::filter(evid == 1) |>
    dplyr::select(scenario, id, time, amt)
  conc_obj <- PKNCA::PKNCAconc(conc, conc ~ time | scenario + id)
  dose_obj <- PKNCA::PKNCAdose(dose, amt ~ time | scenario + id)
  ints <- data.frame(start = 648, end = 672, auclast = TRUE, cmax = TRUE)
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = ints))
  as.data.frame(res$result) |>
    dplyr::select(scenario, PPTESTCD, PPORRES) |>
    tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
}

quiz <- nca_forest(sim_forest, "Cc")
ac886 <- nca_forest(sim_forest, "Cc_ac886")

ratios <- quiz |>
  dplyr::rename(auc_q = auclast, cmax_q = cmax) |>
  dplyr::left_join(ac886 |> dplyr::rename(auc_m = auclast, cmax_m = cmax), by = "scenario") |>
  dplyr::mutate(auc_sum = auc_q + auc_m)
ref_row <- ratios[ratios$scenario == "Reference", ]
ratios <- ratios |>
  dplyr::mutate(
    q_auc = auc_q / ref_row$auc_q, q_cmax = cmax_q / ref_row$cmax_q,
    m_auc = auc_m / ref_row$auc_m, m_cmax = cmax_m / ref_row$cmax_m,
    sum_auc = auc_sum / ref_row$auc_sum
  )

ratios |>
  dplyr::filter(scenario != "Reference") |>
  dplyr::select(scenario, q_auc, q_cmax, m_auc, m_cmax, sum_auc) |>
  dplyr::mutate(dplyr::across(-scenario, ~ round(.x, 3))) |>
  dplyr::rename(
    "Scenario" = scenario,
    "Quizartinib AUC0-24,ss ratio" = q_auc,
    "Quizartinib Cmax,ss ratio" = q_cmax,
    "AC886 AUC0-24,ss ratio" = m_auc,
    "AC886 Cmax,ss ratio" = m_cmax,
    "Quizartinib + AC886 AUC ratio" = sum_auc
  ) |>
  knitr::kable(caption = "Replicates Figure 4 of Kang 2020 (typical-value ratios vs the reference AML patient, day 28, 60 mg once daily).")
```

The text of Kang 2020 states the published values that can be compared
numerically: strong CYP3A inhibitors increase quizartinib `AUC0-24,ss` by 82%
and `Cmax,ss` by 72%; Black or African American participants have
approximately 35% lower AC886 exposure; every other covariate changes
quizartinib exposure by less than 20%; and the extreme BSA values move AC886
exposure by more than 20% (Figure 4B shows about 1.48 at 1.5 m^2 and 0.72 at
2.3 m^2). For the quizartinib + AC886 sum (Figure 4C), only the strong CYP3A
inhibitor falls outside 0.8-1.25 (about 1.6 in the figure).

```{r forest-compare}
get_r <- function(s, col) ratios[[col]][ratios$scenario == s]
cmp <- data.frame(
  Quantity = c(
    "Strong CYP3A inhibitor: quizartinib AUC ratio",
    "Strong CYP3A inhibitor: quizartinib Cmax ratio",
    "Black race: AC886 AUC ratio"
  ),
  Published = c(1.82, 1.72, 0.65),
  Simulated = c(
    get_r("Strong CYP3A inhibitor", "q_auc"),
    get_r("Strong CYP3A inhibitor", "q_cmax"),
    get_r("Black race", "m_auc")
  )
) |>
  dplyr::mutate(`Difference (%)` = round(100 * (Simulated / Published - 1), 1), Simulated = round(Simulated, 3))
knitr::kable(cmp, caption = "Published covariate effects (Kang 2020 Results and Conclusions) vs simulated.")

other_q <- ratios |>
  dplyr::filter(!scenario %in% c("Reference", "Strong CYP3A inhibitor"))
stopifnot(
  abs(get_r("Strong CYP3A inhibitor", "q_auc") / 1.82 - 1) < 0.05,
  abs(get_r("Strong CYP3A inhibitor", "q_cmax") / 1.72 - 1) < 0.05,
  abs(get_r("Black race", "m_auc") - 0.65) < 0.05,
  all(other_q$q_auc > 0.8 & other_q$q_auc < 1.25),
  all(other_q$q_cmax > 0.8 & other_q$q_cmax < 1.25),
  get_r("BSA 1.5 m^2", "m_auc") > 1.25,
  get_r("BSA 2.3 m^2", "m_auc") < 0.8,
  abs(get_r("Strong CYP3A inhibitor", "sum_auc") - 1.6) < 0.1,
  all(other_q$sum_auc > 0.8 & other_q$sum_auc < 1.25)
)
```

# Analytic steady state

At true steady state the dose-interval AUC of a linear model equals
`F * Dose / CL` for the parent and `fMET * F * Dose / CLm` for the metabolite,
independent of the absorption, distribution and lag parameters. This checks
the bioavailability equation, the clearance brackets and the metabolite
formation term exactly. Here the reference patient and the strong CYP3A
inhibitor patient are dosed for 150 days.

```{r ss-analytic}
ss_cohort <- scen |>
  dplyr::filter(scenario %in% c("Reference", "Strong CYP3A inhibitor", "Black race"))
ev_ss <- make_events(ss_cohort, dose = 60, n_doses = 150, obs_times = seq(149 * 24, 150 * 24, by = 0.05))
# 150 days with no observations before the last one exceeds the default
# solver step budget, so maxsteps is raised.
sim_ss <- rxode2::rxSolve(
  rxode2::zeroRe(mod), ev_ss,
  keep = c("scenario"), returnType = "data.frame", useLinCmt = FALSE,
  maxsteps = 1e6
)
stopifnot(!anyNA(sim_ss$Cc), !anyNA(sim_ss$Cc_ac886))
trap <- function(t, y) sum(diff(t) * (head(y, -1) + tail(y, -1)) / 2)
auc_ss <- sim_ss |>
  dplyr::group_by(scenario) |>
  dplyr::summarise(
    auc_q = trap(time, Cc), auc_m = trap(time, Cc_ac886),
    f = fdepot[1], cl = cl[1], cl_m = cl_ac886[1], .groups = "drop"
  ) |>
  dplyr::mutate(
    auc_q_expected = 1000 * f * 60 / cl,
    auc_m_expected = 1000 * 0.5 * f * 60 / cl_m,
    q_err_pct = 100 * (auc_q / auc_q_expected - 1),
    m_err_pct = 100 * (auc_m / auc_m_expected - 1)
  )
knitr::kable(auc_ss |> dplyr::select(scenario, auc_q, auc_q_expected, q_err_pct, auc_m, auc_m_expected, m_err_pct), digits = 2)
stopifnot(all(abs(auc_ss$q_err_pct) < 0.5), all(abs(auc_ss$m_err_pct) < 0.5))
```

The analytic steady-state strong-CYP3A-inhibitor ratio for quizartinib is
`1.136 * 1.082 / 0.641 = 1.92`, higher than the published 1.82: quizartinib
accumulates slowly through its deep peripheral compartment, and the slower
clearance with an inhibitor lengthens the approach to steady state further, so
the ratio still rises between day 28 (`r round(get_r("Strong CYP3A inhibitor", "q_auc"), 2)`)
and full steady state. The day-28 value matches the published ratio, which
suggests the Figure 4 simulations were evaluated after one 28-day cycle.

# Typical profiles and accumulation

Kang 2020 reports observed Tmax of about 2 hours in patients and 4 hours in
healthy volunteers (Discussion), and median accumulation ratios of 4.8 for
quizartinib and 8.4 for AC886 in QuANTUM-R patients (Results "Data").

```{r typical-profiles}
tv_cohort <- dplyr::bind_rows(
  ref_patient |> dplyr::mutate(scenario = "AML patient, 60 mg once daily"),
  ref_patient |> dplyr::mutate(scenario = "Healthy volunteer, 30 mg single dose", DIS_AML = 0, ALB = 44, AGE = 33)
) |>
  dplyr::mutate(id = dplyr::row_number())
ev_tv <- dplyr::bind_rows(
  make_events(tv_cohort[1, ], dose = 60, n_doses = 15, obs_times = seq(0, 15 * 24, by = 0.1)),
  make_events(tv_cohort[2, ], dose = 30, n_doses = 1, obs_times = seq(0, 504, by = 0.1))
)
sim_tv <- rxode2::rxSolve(
  rxode2::zeroRe(mod), ev_tv,
  keep = c("scenario"), returnType = "data.frame", useLinCmt = FALSE
)

sim_tv |>
  dplyr::filter(time <= 72) |>
  tidyr::pivot_longer(c(Cc, Cc_ac886), names_to = "analyte", values_to = "conc") |>
  dplyr::mutate(analyte = ifelse(analyte == "Cc", "Quizartinib", "AC886")) |>
  ggplot(aes(time, conc, colour = analyte)) +
  geom_line() +
  facet_wrap(~scenario, scales = "free_y") +
  labs(x = "Time (h)", y = "Concentration (ng/mL)", colour = NULL,
       title = "Typical-value profiles over the first 72 hours")

aml <- sim_tv |> dplyr::filter(id == 1)
hv <- sim_tv |> dplyr::filter(id == 2)
day_auc <- function(d, day, col) {
  x <- d[d$time >= (day - 1) * 24 & d$time <= day * 24, ]
  trap(x$time, x[[col]])
}
acc <- data.frame(
  Quantity = c("Tmax, AML day 1 (h)", "Tmax, healthy volunteer (h)",
               "Accumulation ratio quizartinib (day 15 / day 1 AUC0-24)",
               "Accumulation ratio AC886 (day 15 / day 1 AUC0-24)"),
  Published = c(2, 4, 4.8, 8.4),
  Simulated = c(
    aml$time[aml$time <= 24][which.max(aml$Cc[aml$time <= 24])],
    hv$time[which.max(hv$Cc)],
    day_auc(aml, 15, "Cc") / day_auc(aml, 1, "Cc"),
    day_auc(aml, 15, "Cc_ac886") / day_auc(aml, 1, "Cc_ac886")
  )
)
acc$Simulated <- round(acc$Simulated, 2)
knitr::kable(acc, caption = "Typical-value absorption and accumulation vs the observed values quoted by Kang 2020.")
stopifnot(
  acc$Simulated[1] >= 1.5 && acc$Simulated[1] <= 3,
  acc$Simulated[2] >= 2.5 && acc$Simulated[2] <= 5,
  abs(acc$Simulated[3] / 4.8 - 1) < 0.2,
  abs(acc$Simulated[4] / 8.4 - 1) < 0.2
)
```

The published Tmax values and accumulation ratios are observed medians rather
than model outputs, so only agreement within about 20% is expected; the model
reproduces the faster absorption in patients (ka 1.68 vs 0.874 1/h).

# Virtual cohorts and stochastic simulation

Two cohorts of 200 are simulated with full interindividual, interoccasion
(occasion 1 throughout) and residual variability: healthy volunteers given a
single fasted 30 mg dose (the design of AC220-015/-016/-018) and AML patients
given 60 mg once daily for 15 days (the QuANTUM-R target dose). Covariates are
drawn around the Table 2 medians, with out-of-range draws redrawn rather than
truncated.

```{r cohort}
rxode2::rxSetSeed(20201201)
draw_in_range <- function(n, median, sdlog, lo, hi) {
  x <- rlnorm(n, log(median), sdlog)
  bad <- x < lo | x > hi
  while (any(bad)) {
    x[bad] <- rlnorm(sum(bad), log(median), sdlog)
    bad <- x < lo | x > hi
  }
  x
}
n <- 200
hv_cohort <- data.frame(
  id = seq_len(n), group = "Healthy volunteers, 30 mg single dose",
  DIS_AML = 0, CONMED_CYP3A4_INH_STRONG = 0,
  ALB = 10 * draw_in_range(n, 4.4, 0.07, 3.3, 5.2),
  BSA = draw_in_range(n, 1.9, 0.10, 1.4, 2.5),
  AGE = draw_in_range(n, 33, 0.30, 18, 66),
  RACE_BLACK = rbinom(n, 1, 0.323), FED = 0, STUDY_AC220019 = 0, OCC = 1
)
aml_cohort <- data.frame(
  id = n + seq_len(n), group = "AML patients, 60 mg once daily",
  DIS_AML = 1, CONMED_CYP3A4_INH_STRONG = 0,
  ALB = 10 * draw_in_range(n, 3.7, 0.12, 2.1, 4.8),
  BSA = draw_in_range(n, 1.9, 0.12, 1.3, 2.8),
  AGE = draw_in_range(n, 55, 0.25, 19, 81),
  RACE_BLACK = rbinom(n, 1, 0.037), FED = 0, STUDY_AC220019 = 0, OCC = 1
)
ev_hv <- make_events(hv_cohort, dose = 30, n_doses = 1,
                     obs_times = sort(unique(c(seq(0, 24, by = 0.25), seq(25, 504, by = 1)))))
ev_aml <- make_events(aml_cohort, dose = 60, n_doses = 15,
                      obs_times = c(seq(0, 24, by = 0.25), seq(14 * 24, 15 * 24, by = 0.25)))
sim_hv <- rxode2::rxSolve(mod, ev_hv, keep = c("group"), returnType = "data.frame", useLinCmt = FALSE)
sim_aml <- rxode2::rxSolve(mod, ev_aml, keep = c("group"), returnType = "data.frame", useLinCmt = FALSE)
```

```{r vpc-plot}
pct <- function(d) {
  d |>
    dplyr::group_by(group, time) |>
    dplyr::summarise(
      p05 = quantile(Cc, 0.05), p50 = median(Cc), p95 = quantile(Cc, 0.95),
      .groups = "drop"
    )
}
dplyr::bind_rows(
  pct(sim_hv),
  pct(sim_aml |> dplyr::filter(time >= 14 * 24) |> dplyr::mutate(time = time - 14 * 24))
) |>
  ggplot(aes(time, p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~group, scales = "free") +
  labs(x = "Time after dose (h; AML: day 15 dosing interval)",
       y = "Quizartinib (ng/mL)",
       title = "Simulated quizartinib: median and 90% interval",
       caption = "Compare with the pcVPC in Figure 3A of Kang 2020.")
```

# PKNCA

```{r pknca}
run_nca <- function(sim, events, conc_col, intervals) {
  conc <- sim |>
    dplyr::filter(!is.na(.data[[conc_col]])) |>
    dplyr::transmute(group, id, time, conc = .data[[conc_col]])
  dose <- events |>
    dplyr::filter(evid == 1) |>
    dplyr::select(group, id, time, amt)
  conc_obj <- PKNCA::PKNCAconc(conc, conc ~ time | group + id)
  dose_obj <- PKNCA::PKNCAdose(dose, amt ~ time | group + id)
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}
int_hv <- data.frame(start = 0, end = 504, cmax = TRUE, tmax = TRUE, auclast = TRUE)
int_aml <- data.frame(
  start = c(0, 336), end = c(24, 360),
  cmax = TRUE, tmax = TRUE, auclast = TRUE
)
nca_q <- dplyr::bind_rows(
  as.data.frame(run_nca(sim_hv, ev_hv, "Cc", int_hv)$result),
  as.data.frame(run_nca(sim_aml, ev_aml, "Cc", int_aml)$result)
)
nca_m <- dplyr::bind_rows(
  as.data.frame(run_nca(sim_hv, ev_hv, "Cc_ac886", int_hv)$result),
  as.data.frame(run_nca(sim_aml, ev_aml, "Cc_ac886", int_aml)$result)
)
summ <- function(res, analyte) {
  res |>
    dplyr::group_by(group, start, end, PPTESTCD) |>
    dplyr::summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
    tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
    dplyr::mutate(analyte = analyte)
}
nca_tab <- dplyr::bind_rows(summ(nca_q, "Quizartinib"), summ(nca_m, "AC886"))
nca_tab |>
  dplyr::select(analyte, group, start, end, cmax, tmax, auclast) |>
  dplyr::mutate(dplyr::across(c(cmax, tmax, auclast), ~ signif(.x, 3))) |>
  dplyr::rename(
    "Analyte" = analyte, "Group" = group, "Interval start (h)" = start,
    "Interval end (h)" = end, "Median Cmax (ng/mL)" = cmax,
    "Median Tmax (h)" = tmax, "Median AUClast (ng*h/mL)" = auclast
  ) |>
  knitr::kable(caption = "Median simulated NCA parameters by group and interval.")

tmax_q <- nca_q |>
  dplyr::filter(PPTESTCD == "tmax", start %in% c(0)) |>
  dplyr::group_by(group) |>
  dplyr::summarise(tmax = median(PPORRES), .groups = "drop")
stopifnot(
  tmax_q$tmax[tmax_q$group == "AML patients, 60 mg once daily"] < tmax_q$tmax[tmax_q$group == "Healthy volunteers, 30 mg single dose"]
)
```

Kang 2020 does not report NCA parameters for the simulated or observed
populations, so there is no published NCA table to compare against; the
covariate-ratio and analytic steady-state checks above serve as the numeric
validation, and the median Tmax values reproduce the faster absorption in
patients described in the Discussion.

# Assumptions and deviations

* **Dose amount.** Kang 2020 states every dose as mg of quizartinib
  dihydrochloride (20, 30, 60 and 90 mg, "equivalent to 17.7, 26.5, 53.0 and
  79.5 mg free base") and does not say which amount entered the dataset. The
  model is used with the salt-form dose, the paper's own dose nomenclature
  throughout; if the analysis dataset used the free-base amount, predicted
  concentrations would be 13% higher (30/26.5). The covariate ratios and
  accumulation checks above are independent of this choice.
* **IIV and IOV scale.** Tables 3 and 4 print each random effect as "% CV"
  without a formula. The paper converts its log-scale residual variances to
  %CV as 100 x sqrt(variance) (Table 4 footnote a: 0.0690 -> 26.3%, 0.169 ->
  41.2%), and other analyses by the same analysis group in this library use
  the same convention, so each variance is taken as (CV/100)^2. If the
  published %CV were instead exact log-normal CVs, the variances would be
  smaller (for example 0.862 rather than 1.369 for the 117% CV on AC886 Vc).
* **Occasions.** Table 3 reports IOV on occasions 1-4 for CL and Vc and on
  occasions 1 and 3 only for F1, without defining the occasions. The model
  exposes an `OCC` column (1-4) and applies the F1 IOV only on occasions 1 and
  3, as printed; `OCC = 1` is used throughout the simulations here.
* **F1 and ka variability.** The single F1 IIV (34.8% CV, printed on the
  "relative F1 for patients" row) and the single ka IIV (38.5%, printed on the
  healthy-volunteer ka row) are applied to all participants, as no separate
  variance is reported for the other population.
* **AC886 formation.** Figure 2 shows AC886 formed at `fMET * CL` from the
  quizartinib central compartment with no molecular-weight conversion, and the
  paper reports none; the model follows the figure. fMET is fixed at 0.5 and is
  not identifiable (only oral data), so AC886 CL and volumes are apparent values
  scaled by it.
* **AC886 central volume covariates.** The Results text says strong CYP3A
  inhibitor use "and patient status" were significant on the AC886 central
  volume, but equation 8 and Table 4 contain only the strong-inhibitor effect,
  and the September 2020 correction to the abstract removed "patient status"
  from the AC886 covariates. The equation and table are followed.
* **Sequential fit.** The AC886 model was estimated with individual quizartinib
  parameters fixed at their post hoc values. This file expresses both moieties
  as one joint model so they can be simulated together.
* **Fed status.** The food effect was estimated only within study AC220-019;
  `FED` should be 0 for all other studies, and `STUDY_AC220019 = 1` switches
  the relative bioavailability to equation 3.
* **Albumin units.** The canonical `ALB` column is in g/L; the model converts
  to g/dL internally so the published centring value (4.1 g/dL) applies.
* **Figure 4 simulation day.** Not stated in the paper; 28 days of dosing
  reproduces the published strong-CYP3A-inhibitor ratios, while the analytic
  steady state gives a higher quizartinib ratio (1.92).
* **Residual error.** Quizartinib uses the combined error
  `Var = F^2 * sigma_prop + sigma_add` with population-specific proportional
  variances and a shared additive variance, as in Table 3 footnotes b and c.
  AC886 uses additive error on log-transformed concentrations with
  population-specific variances.
