---
title: "Radotinib (Kang 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Radotinib (Kang 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
ui <- rxode2::rxode(readModelDb("Kang_2025_radotinib"))
```

* Citation: `r ui$reference`
* Description: `r ui$description`
* Article: <https://doi.org/10.3390/ph18111705>

Radotinib is an oral second-generation BCR-ABL1 tyrosine kinase inhibitor
approved in the Republic of Korea for chronic-phase chronic myeloid leukemia
(CML-CP). Kang 2025 is the first population PK analysis of the drug.

# Population

The model was fit to 640 plasma concentrations from 47 Asian participants
pooled across two clinical studies (Kang 2025 Results section 2.1, Table 1):

* **23 healthy male volunteers** (Korean, ClinicalTrials.gov NCT06461078), each
  given a single oral 400 mg dose under fasting conditions, with 306
  concentrations sampled pre-dose and at 0.5 to 48 h post-dose.
* **24 patients with Ph+ CML-CP** (Chinese, NCT03722420), given 300 mg orally
  twice daily under fasting conditions, with 334 concentrations sampled
  pre-dose and to 12 h post-dose on days 1 and 14.

Pooled baseline characteristics: median (range) age 31 (20-72) years, weight
65.7 (47.0-96.0) kg, height 171 (155-186) cm, BMI 22.8 (18.4-30.3) kg/m^2;
36 males (77%) and 11 females (23%); creatinine clearance 114 (72-200) mL/min.
All healthy volunteers were male; the female participants are all in the
CML-CP arm.

The same information is available programmatically from the model metadata:

```{r population}
str(ui$population, max.level = 1)
```

# Source trace

Each `ini()` entry in `inst/modeldb/specificDrugs/Kang_2025_radotinib.R`
carries an in-file comment naming its source location. They are collected here
for review. Equation numbers refer to Kang 2025 Results section 2.3.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL/F, CML-CP, circadian mesor) | 23.0 L/h | Table 2 row `CL/F (L/h)`, RSE 7.3% |
| `lvc` (Vc/F at age 31) | 383 L | Table 2 row `Vc/F (L)`, RSE 13.8% |
| `lq` (Q/F) | 132 L/h | Table 2 row `Q/F (L/h)`, RSE 11.6% |
| `lvp` (Vp/F) | 519 L | Table 2 row `Vp/F (L)`, RSE 15.6% |
| `lka` (ka) | 1.59 1/h | Table 2 row `ka (h-1)`, RSE 21.6% |
| `lmtt` (MTT) | 1.88 h | Table 2 row `MTT (h)`, RSE 10.4% |
| `lnn` (N) | 6.58 | Table 2 row `N`, RSE 17.0% |
| `amp_cl` (circadian fractional amplitude) | 0.683 | Table 2 row `CL/F Circadian effect`, RSE 14.8%; equation 1 |
| `acrophase_cl` (circadian phase shift) | 7 h | Equation 1 only (`cos{2*pi*(TIME - 7)/24}`); not tabulated, no RSE |
| `e_dis_cml_cl` (disease status on CL/F) | 0.646 | Table 2 row `CL/F Disease status`, RSE 30.0%; equation 1, Table 2 footnote a |
| `e_age_vc` (age on Vc/F) | -0.0129 /year | Table 2 row `Vc/F Age`, RSE 37.6%; equation 2, Table 2 footnote b |
| `etalcl` | 0.389^2 = 0.151321 | Table 2 row `omega CL/F` = 0.389 (a SD), RSE 20.1% |
| `etalmtt` | 0.1 FIXED | Results 2.3 ("variances ... fixed at ... 0.1"); Table 2 `omega MTT` = 0.316 = sqrt(0.1) |
| `etalnn` | 0.2 FIXED | Results 2.3 ("variances ... fixed at ... 0.2"); Table 2 `omega N` = 0.447 = sqrt(0.2) |
| `etaiov_v_1` / `etaiov_v_2` | 0.698^2 = 0.487204 | Table 2 row `omega IOV` = 0.698 (a SD), RSE 25.1%; equations 2 and 4 |
| `propSd` | 0.200 | Table 2 row `sigma proportional (%)` = 20.0 |
| Transit-chain input, `ktr = (N + 1)/MTT` | n/a | Methods 4.3 ("transit compartment model with first-order absorption"); Savic 2007 parameterisation |
| `d/dt(central)`, `d/dt(peripheral1)` | n/a | Results 2.3 ("two-compartment model ... first-order elimination") |
| `Cc ~ prop(propSd)` | n/a | Results 2.3 ("Residual variability was most appropriately described using a proportional model") |

## The omega column is a standard deviation, not a variance

Table 2's interindividual-variability block is headed `omega <parameter>` with
no scale stated. The paper settles it internally: Results section 2.3 says the
IIV **variances** for MTT and N "were fixed at ... 0.1 and 0.2, respectively",
and Table 2 prints `omega MTT = 0.316` and `omega N = 0.447`, which are
`sqrt(0.1)` and `sqrt(0.2)`. Every entry in that column is therefore a standard
deviation and is squared in the model file to reach the log-scale variance.

```{r omega-scale}
c(MTT = sqrt(0.1), N = sqrt(0.2))   # -> 0.3162, 0.4472, as printed in Table 2
```

# Structural checks against published derived quantities

These checks are deterministic functions of the published point estimates -- no
simulated cohort is involved -- so they are asserted tightly. They are the
sharpest available test that the parameter values, the covariate forms and the
disposition structure were transcribed correctly: a mis-read clearance, volume,
or covariate direction moves them immediately.

```{r structural-gates}
th <- ui$theta

# Terminal half-life of the two-compartment disposition model.
terminal_hl <- function(cl, vc, vp, q) {
  k10 <- cl / vc; k12 <- q / vc; k21 <- q / vp
  s <- k10 + k12 + k21
  beta <- 0.5 * (s - sqrt(s^2 - 4 * k21 * k10))
  log(2) / beta
}

cl_cml <- exp(th[["lcl"]])
cl_hv  <- cl_cml * (1 + th[["e_dis_cml_cl"]])   # equation 1 with DIS_CML = 0
vc31   <- exp(th[["lvc"]])
vp     <- exp(th[["lvp"]])
qq     <- exp(th[["lq"]])

vc_at <- function(age) vc31 * (1 + th[["e_age_vc"]] * (age - 31))

checks <- tibble::tribble(
  ~Quantity,                              ~Published, ~Model,
  "CL/F, CML-CP patient (L/h)",                 23.0, cl_cml,
  "CL/F, healthy volunteer (L/h)",              37.9, cl_hv,
  "Terminal half-life, CML-CP (h)",             28.8, terminal_hl(cl_cml, vc31, vp, qq),
  "Terminal half-life, healthy volunteer (h)",  18.2, terminal_hl(cl_hv,  vc31, vp, qq),
  "Vc/F at age 20 (L)",                          437, vc_at(20),
  "Vc/F at age 72 (L)",                          180, vc_at(72),
  "Total V/F = Vc/F + Vp/F (L)",                 902, vc31 + vp
) |>
  mutate(`% diff` = 100 * (Model - Published) / Published)

knitr::kable(checks, digits = c(0, 1, 2, 2),
             caption = "Derived quantities the model must reproduce exactly. Published values from Kang 2025 Abstract, Results section 2.3 and Discussion.")

# Deterministic -- these are closed-form functions of the published thetas, so
# the tolerances are tight on purpose. There is no cohort and no RNG here.
stopifnot(
  abs(cl_cml - 23.0) < 0.01,
  abs(cl_hv  - 37.9) < 0.05,
  abs(terminal_hl(cl_cml, vc31, vp, qq) - 28.8) < 0.1,
  abs(terminal_hl(cl_hv,  vc31, vp, qq) - 18.2) < 0.1,
  abs(vc_at(20) - 437) < 1,
  abs(vc_at(72) - 180) < 1,
  abs((vc31 + vp) - 902) < 1
)
```

The half-lives are the strongest of these: 28.8 h and 18.2 h are reproduced to
better than 0.05 h from the four disposition parameters and the disease
coefficient, which pins `CL/F`, `Vc/F`, `Vp/F`, `Q/F` and the 0.646 disease
effect simultaneously.

## The circadian rhythm on CL/F

Equation 1 modulates apparent clearance with a 24 h cosinor whose amplitude is
**fractional** -- it multiplies the mesor rather than being expressed in L/h:

$$\mathrm{CL}/F_i = 23.0 \times \left[1 + 0.646 (1 - \mathrm{CML}_i)\right] \times \left[1 + 0.683 \cos\!\left(\frac{2\pi(\mathrm{TIME} - 7)}{24}\right)\right] \times e^{\eta_i}$$

`TIME` is the NONMEM time variable, i.e. hours since the first dose, not clock
time. Kang 2025's Discussion states that the model puts peak CL/F activity at
16:00, "in close agreement" with the reported ~15:00 peak of hepatic CYP3A
activity; a 7 h offset reaches 16:00 from a ~09:00 morning dose, which is the
only reading consistent with both equation 1 and that sentence.

```{r circadian}
mod_typ <- rxode2::zeroRe(readModelDb("Kang_2025_radotinib"))

ev_day <- data.frame(
  id = 1L, time = c(0, seq(0, 48, by = 0.25)),
  evid = c(1L, rep(0L, 193)), amt = c(300, rep(NA_real_, 193)),
  cmt = c("depot", rep("central", 193)),
  DIS_CML = 1, AGE = 32, OCC = 2L
)
day <- rxode2::rxSolve(mod_typ, ev_day, returnType = "data.frame")

cl_range <- range(day$cl)
# Average over a HALF-OPEN 24 h period: including both t = 24 and t = 48 would
# count the same phase twice and bias the mean off the mesor.
cl_mesor <- mean(day$cl[day$time >= 24 & day$time < 48])

tibble::tibble(
  Quantity = c("CL/F mesor over a full 24 h cycle (L/h)",
               "CL/F minimum (L/h)", "CL/F maximum (L/h)",
               "Hours after dose at which CL/F peaks"),
  Model = c(cl_mesor, cl_range[1], cl_range[2],
            day$time[which.max(day$cl)]),
  Expected = c(23.0, 23.0 * (1 - 0.683), 23.0 * (1 + 0.683), 7)
) |>
  knitr::kable(digits = 2,
               caption = "The cosinor term as encoded, against equation 1 evaluated by hand.")

stopifnot(
  abs(cl_mesor - 23.0) < 0.01,
  abs(cl_range[1] - 23.0 * (1 - 0.683)) < 0.01,
  abs(cl_range[2] - 23.0 * (1 + 0.683)) < 0.01,
  abs(day$time[which.max(day$cl)] - 7) < 0.3
)
```

Because the amplitude 0.683 is smaller than 1, the bracket never changes sign
and CL/F stays positive across the whole cycle, ranging from 7.3 to 38.7 L/h in
a CML-CP patient.

## Mass balance across a steady-state day

At steady state the drug eliminated over one full 24 h circadian cycle must
equal the dose given over that cycle. With a time-varying clearance this is a
non-trivial check of the ODE system, the transit input and the cosinor term
together -- it fails if the dose is delivered twice (the classic transit-model
bug where the depot bolus is not suppressed), if the input kernel is
mis-normalised, or if the circadian term is applied to the wrong quantity.

```{r mass-balance}
ss_profile <- function(dose, ii, ndays = 15, by = 0.05) {
  ndose <- ceiling(ndays * 24 / ii)
  tobs  <- seq((ndays - 1) * 24, ndays * 24, by = by)
  data.frame(
    id = 1L,
    time = c(seq(0, by = ii, length.out = ndose), tobs),
    evid = c(rep(1L, ndose), rep(0L, length(tobs))),
    amt  = c(rep(dose, ndose), rep(NA_real_, length(tobs))),
    cmt  = c(rep("depot", ndose), rep("central", length(tobs))),
    DIS_CML = 1, AGE = 32, OCC = 2L
  ) |>
    (\(ev) rxode2::rxSolve(mod_typ, ev, returnType = "data.frame"))() |>
    dplyr::filter(!is.na(Cc))
}

trapz <- function(x, y) sum(diff(x) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)

mb <- lapply(
  list(c(300, 12), c(300, 24), c(400, 24), c(500, 24), c(600, 24)),
  function(p) {
    s <- ss_profile(p[1], p[2])
    # Eliminated amount over the day = integral of CL(t) * C(t) dt.
    # Cc is ng/mL = ug/L, so divide by 1000 to reach mg/L before multiplying
    # by CL/F in L/h.
    eliminated <- trapz(s$time, s$cl * s$Cc / 1000)
    daily <- p[1] * 24 / p[2]
    tibble::tibble(Regimen = sprintf("%d mg %s", p[1], ifelse(p[2] == 12, "BID", "QD")),
                   `Dose over 24 h (mg)` = daily,
                   `Eliminated over 24 h (mg)` = eliminated,
                   `Ratio` = eliminated / daily)
  }
) |> bind_rows()

knitr::kable(mb, digits = c(0, 0, 2, 4),
             caption = "Steady-state mass balance: amount eliminated over one circadian cycle versus the dose given.")

# Deterministic (typical-value solve, no RNG): the residual is numerical
# integration error on a 0.02 h grid, so a tight bound is correct here.
stopifnot(all(abs(mb$Ratio - 1) < 0.005))
```

# Virtual cohort

The observed concentrations are not public, so the sections below simulate
virtual cohorts whose covariate distributions approximate the published trial
demographics (Table 1).

```{r cohort}
# set.seed() seeds R's RNG for the covariate draws below. It does NOT seed
# rxode2's simulation RNG, whose streams are partitioned per solver thread --
# so the etas drawn here differ between a 16-thread workstation and a 2-core CI
# runner and no seed can make them agree. Every assertion downstream is written
# to hold for any cohort this model can produce.
set.seed(20251110)
n_arm <- 150L

# Age draws: a truncated normal matched to the per-arm median and range of
# Kang 2025 Table 1 (CML-CP median 32, range 21-72; healthy volunteers median
# 29, range 20-51). The paper reports only median and range, so the shape of
# the distribution between those points is an assumption (see Assumptions).
draw_age <- function(n, med, lo, hi, sd) {
  a <- round(rnorm(n, med, sd))
  pmin(pmax(a, lo), hi)
}

# The five CML-CP regimens are compared against each other and against Table 3,
# so they use COMMON RANDOM NUMBERS: the same n_arm subjects (same ages, same
# ids, and -- because rxode2's eta stream is keyed by id and the seed is reset
# before each solve -- the same random effects) are put through every regimen.
# Independent draws per arm would add roughly a 4% standard error to each arm's
# median on top of the systematic offset discussed below, which is enough to
# make the cross-regimen comparison look non-dose-proportional purely by chance.
age_cml <- draw_age(n_arm, 32, 21, 72, 12)
age_hv  <- draw_age(n_arm, 29, 20, 51, 8)

make_arm <- function(dose, ii, is_cml, ndays = 21, obs_by = 0.25) {
  if (is_cml) {
    ndose  <- ceiling(ndays * 24 / ii)
    dose_t <- seq(0, by = ii, length.out = ndose)
    obs_t  <- seq((ndays - 1) * 24, ndays * 24, by = obs_by)
    subj   <- tibble::tibble(id = seq_len(n_arm), AGE = age_cml)
    occ    <- 2L
  } else {
    dose_t <- 0
    obs_t  <- seq(0, 48, by = obs_by)
    subj   <- tibble::tibble(id = seq_len(n_arm), AGE = age_hv)
    occ    <- 1L
  }
  bind_rows(
    tidyr::crossing(subj, time = dose_t) |>
      mutate(evid = 1L, amt = dose, cmt = "depot"),
    tidyr::crossing(subj, time = obs_t) |>
      mutate(evid = 0L, amt = NA_real_, cmt = "central")
  ) |>
    mutate(DIS_CML = as.integer(is_cml), OCC = occ) |>
    arrange(id, time, desc(evid))
}

arms <- list(
  "300 mg BID"             = list(dose = 300, ii = 12, cml = TRUE),
  "300 mg QD"              = list(dose = 300, ii = 24, cml = TRUE),
  "400 mg QD"              = list(dose = 400, ii = 24, cml = TRUE),
  "500 mg QD"              = list(dose = 500, ii = 24, cml = TRUE),
  "600 mg QD"              = list(dose = 600, ii = 24, cml = TRUE),
  "HV 400 mg single dose"  = list(dose = 400, ii = NA, cml = FALSE)
)

events <- lapply(names(arms), function(nm) {
  a <- arms[[nm]]
  make_arm(a$dose, a$ii, a$cml) |> mutate(regimen = nm)
}) |> bind_rows()

# Within an arm the (id, time, evid) key must be unique; ids intentionally
# REPEAT across arms (that is what makes the random numbers common), and each
# arm is solved in its own rxSolve call so they never share a subject key.
stopifnot(!anyDuplicated(unique(events[, c("regimen", "id", "time", "evid")])))
stopifnot(length(unique(events$regimen)) == 6L)
```

# Simulation

```{r simulate}
mod <- readModelDb("Kang_2025_radotinib")

# One solve per arm, with the seed reset before each, so every arm draws the
# same eta sequence for subject i. Note rxSetSeed() fixes the stream per solver
# thread, not across thread counts -- this makes the arms comparable to each
# other on any machine, but does not make the cohort identical between machines.
sim <- lapply(names(arms), function(nm) {
  ev <- events |> filter(regimen == nm) |> select(-regimen)
  rxode2::rxSetSeed(20251110)
  rxode2::rxSolve(mod, events = ev, keep = c("AGE", "DIS_CML"),
                  returnType = "data.frame") |>
    mutate(regimen = nm)
}) |> bind_rows() |> filter(!is.na(Cc))

stopifnot(nrow(sim) > 0, !anyNA(sim$Cc), all(sim$Cc >= 0))
```

# Replicate published figures

## Figure 3 -- simulated steady-state concentration-time profiles

Kang 2025 Figure 3 shows the median and 90% prediction interval of simulated
steady-state profiles for 300 mg BID and 300, 400, 500 and 600 mg QD in virtual
CML-CP patients. Panel F overlays all five.

```{r figure-3, fig.height = 6}
ss <- sim |>
  filter(DIS_CML == 1) |>
  mutate(tau_time = time - 20 * 24)   # hours into the final 24 h window

ss |>
  group_by(regimen, tau_time) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = median(Cc),
            Q95 = quantile(Cc, 0.95), .groups = "drop") |>
  ggplot(aes(tau_time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue4", linewidth = 0.7) +
  facet_wrap(~regimen) +
  scale_x_continuous(breaks = seq(0, 24, by = 6)) +
  labs(x = "Time into the steady-state dosing day (h)",
       y = "Radotinib plasma concentration (ng/mL)",
       title = "Steady-state profiles in virtual CML-CP patients",
       caption = "Replicates Figure 3 of Kang 2025 (median with 90% prediction interval).")
```

Kang 2025 Results section 2.4 describes these profiles as rising "with a delay
of approximately 0.5 h", reaching Cmax "in approximately 3 h" and then declining
until the next dose.

```{r figure-3-shape}
shape <- ss |>
  group_by(regimen, tau_time) |>
  summarise(Q50 = median(Cc), .groups = "drop") |>
  group_by(regimen) |>
  summarise(tmax_median_profile = tau_time[which.max(Q50)], .groups = "drop")

knitr::kable(shape, digits = 2,
             caption = "Time of the peak of the median profile within the steady-state day (paper: approximately 3 h).")

# Cohort-derived, so the bound is a band around the paper's stated ~3 h rather
# than an exact match. Realised 3.00-3.25 h across all five arms locally.
stopifnot(all(shape$tmax_median_profile >= 2), all(shape$tmax_median_profile <= 5))
```

## Healthy volunteers versus patients

The paper's headline covariate finding is that CL/F is 39.2% slower in CML-CP
patients than in healthy volunteers, lengthening the terminal half-life from
18.2 h to 28.8 h.

```{r hv-vs-cml}
sim |>
  filter(regimen == "HV 400 mg single dose") |>
  group_by(time) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = median(Cc),
            Q95 = quantile(Cc, 0.95), .groups = "drop") |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "darkorange") +
  geom_line(colour = "darkorange3", linewidth = 0.7) +
  scale_y_log10() +
  labs(x = "Time after dose (h)", y = "Radotinib plasma concentration (ng/mL)",
       title = "Healthy volunteers, single oral 400 mg dose",
       caption = "Sampling schedule of NCT06461078 (Kang 2025 Methods section 4.1).")
```

# PKNCA validation

NCA is computed with PKNCA over the final 24 h dosing day for the CML-CP arms
(the window Table 3 reports) and over 0-48 h for the single-dose healthy-
volunteer arm.

```{r pknca-ss}
# Shift the steady-state window onto a 0-24 h axis for the NCA intervals. The
# SOLVE must stay on the absolute time axis because the circadian term depends
# on it; only the reported times are shifted here.
t0_ss <- 20 * 24

nca_conc <- sim |>
  filter(DIS_CML == 1) |>
  transmute(id, regimen, time = time - t0_ss, Cc)

nca_dose <- events |>
  filter(DIS_CML == 1, evid == 1, time >= t0_ss, time < t0_ss + 24) |>
  transmute(id, regimen, time = time - t0_ss, amt)

# Every arm must have a dose at the start of the window and an observation at
# both ends of it, or PKNCA silently anchors AUC somewhere else.
stopifnot(
  all(tapply(nca_dose$time, nca_dose$regimen, min) == 0),
  all(tapply(nca_conc$time, nca_conc$regimen, min) == 0),
  all(tapply(nca_conc$time, nca_conc$regimen, max) == 24)
)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(nca_conc), Cc ~ time | regimen + id,
                             concu = "ng/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(as.data.frame(nca_dose), amt ~ time | regimen + id,
                             doseu = "mg")

intervals_ss <- data.frame(
  start = 0, end = 24,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, clast.obs = TRUE, cav = TRUE
)

nca_ss <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                         intervals = intervals_ss))
```

`clast.obs` over the 0-24 h window is the concentration at the end of the
dosing day, i.e. the pre-dose trough -- the quantity Table 3 reports as
`Ctrough`. PKNCA's `cmin` over a single interval would instead return the
minimum *within* the window, which for the BID arm is the 12 h trough.

```{r pknca-ss-summary}
ss_res <- as.data.frame(nca_ss$result) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "clast.obs")) |>
  group_by(regimen, PPTESTCD) |>
  summarise(median = median(PPORRES),
            q25 = quantile(PPORRES, 0.25),
            q75 = quantile(PPORRES, 0.75), .groups = "drop")

ss_res |>
  mutate(Parameter = nlmixr2lib::ncaParamLabel(PPTESTCD)) |>
  select(Regimen = regimen, Parameter, Median = median, `Q1` = q25, `Q3` = q75) |>
  knitr::kable(digits = 1,
               caption = "Simulated steady-state NCA, median (interquartile range) over 150 virtual CML-CP patients per arm.")
```

## Comparison against Kang 2025 Table 3

```{r nca-compare}
published <- tibble::tribble(
  ~regimen,      ~cmax, ~tmax, ~auclast, ~clast.obs,
  "300 mg BID",  1551,      3,    29669,        960,
  "300 mg QD",    955,      3,    14276,        443,
  "400 mg QD",   1274,      3,    19034,        591,
  "500 mg QD",   1592,      3,    23793,        739,
  "600 mg QD",   1910,      3,    28551,        887
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_ss,
  reference = published,
  by        = "regimen",
  units     = c(cmax = "ng/mL", tmax = "h",
                auclast = "ng*h/mL", clast.obs = "ng/mL"),
  tolerance_pct = 20
)

knitr::kable(cmp, digits = 1,
             caption = "Simulated versus Kang 2025 Table 3 (median of 2400 virtual patients). * differs by more than 20%.")
attr(cmp, "footnote")
```

Because the five arms share subjects (common random numbers), the four
once-daily rows are exactly dose-proportional and their offsets against Table 3
are identical to the last digit -- `-11.9%` on AUC, `-9.7%` on Cmax and `-5.6%`
on Ctrough for every one of 300, 400, 500 and 600 mg QD. The twice-daily arm is
offset by `-8.9%` on AUC. The simulated exposures therefore track Table 3 in
dose-proportionality, in profile shape and in `tmax`, but sit systematically
**below** the published medians by roughly a tenth. This is a reproducible
disagreement with the published simulation table rather than a transcription
error in the model, and the next section quantifies why.

Two rows deserve a note. `Tmax` differs by up to 16.7% only because the
observation grid has 0.25 h resolution, so the simulated peak lands at 3.25 or
3.5 h against the paper's rounded `3`; both sit inside Table 3's own reported
interquartile range of 3-4 h for the once-daily regimens. And `Ctrough` for the
twice-daily arm is the one row that comes out *above* the paper (+14.5%),
because with a 24 h rhythm and a 12 h interval the morning and evening troughs
are not equal -- this model puts them at roughly 870 and 1100 ng/mL, and Table 3
reports a single value of 960 between them.

```{r nca-compare-gate}
pct <- suppressWarnings(as.numeric(gsub("[^0-9.+-]", "", cmp[["% diff"]])))
worst <- max(abs(pct), na.rm = TRUE)
worst

# The Table 3 gap is a documented deviation (see below): a uniform ~12% on AUC
# with the tmax grid artifact the largest single row. Realised worst-case
# 16.7%; the bound leaves room for the cohort draw to move but still goes red
# on a mis-transcribed dose, clearance or unit, any of which moves exposure by
# tens of percent.
stopifnot(worst < 30)
```

## Why the simulated exposures sit below Table 3

For a linear model the steady-state AUC over a full 24 h cycle is exactly
`daily dose / CL_effective`, so each row of Table 3 pins the effective apparent
clearance the paper's own simulation must have used.

```{r table3-implied-cl}
implied <- published |>
  mutate(`Daily dose (mg)` = c(600, 300, 400, 500, 600),
         `Implied CL/F (L/h)` = `Daily dose (mg)` / auclast * 1000,
         `Ratio to the 23.0 L/h mesor` = 23.0 / `Implied CL/F (L/h)`) |>
  select(Regimen = regimen, `Daily dose (mg)`, `AUC0-24h (ng*h/mL)` = auclast,
         `Implied CL/F (L/h)`, `Ratio to the 23.0 L/h mesor`)

knitr::kable(implied, digits = c(0, 0, 0, 2, 4),
             caption = "Effective CL/F implied by each row of Kang 2025 Table 3.")
```

The four once-daily rows imply an effective CL/F of 21.0 L/h to four
significant figures -- Table 3 is internally perfectly linear -- and the
twice-daily row implies 20.2 L/h. Both are below the 23.0 L/h mesor of
equation 1.

A time-varying clearance *can* push effective clearance away from the mesor,
because over a steady-state cycle `daily dose = integral of CL(t) C(t) dt`: if
concentration is systematically high when clearance is low, effective clearance
falls below the mesor and AUC rises above `dose / mesor`. Whether that happens,
and by how much, depends on the phase relationship between the dosing schedule
and the rhythm. For a **once-daily** regimen the two are locked, so the effect
is large; for a **twice-daily** regimen the profile repeats twice per rhythm
cycle and the effect largely averages out.

```{r phase-sensitivity}
auc24_at <- function(dose, ii, acro) {
  ndays <- 15; ndose <- ceiling(ndays * 24 / ii)
  tobs <- seq((ndays - 1) * 24, ndays * 24, by = 0.05)
  ev <- data.frame(
    id = 1L,
    time = c(seq(0, by = ii, length.out = ndose), tobs),
    evid = c(rep(1L, ndose), rep(0L, length(tobs))),
    amt  = c(rep(dose, ndose), rep(NA_real_, length(tobs))),
    cmt  = c(rep("depot", ndose), rep("central", length(tobs))),
    DIS_CML = 1, AGE = 32, OCC = 2L
  )
  s <- rxode2::rxSolve(mod_typ, ev, params = c(acrophase_cl = acro),
                       returnType = "data.frame") |> filter(!is.na(Cc))
  trapz(s$time, s$Cc)
}

# Odd-numbered grid so the acrophase of 7 h printed in equation 1 is on it.
acro_grid <- seq(1, 23, by = 2)
phase <- tibble::tibble(
  acrophase = acro_grid,
  `400 mg QD` = vapply(acro_grid, \(a) auc24_at(400, 24, a) / (400 / 23 * 1000), 1),
  `300 mg BID` = vapply(acro_grid, \(a) auc24_at(300, 12, a) / (600 / 23 * 1000), 1)
)

phase |>
  pivot_longer(-acrophase, names_to = "Regimen", values_to = "ratio") |>
  ggplot(aes(acrophase, ratio, colour = Regimen)) +
  geom_hline(yintercept = 19034 / (400 / 23 * 1000), linetype = "dashed",
             colour = "grey30") +
  geom_hline(yintercept = 29669 / (600 / 23 * 1000), linetype = "dotted",
             colour = "grey30") +
  geom_line(linewidth = 0.7) + geom_point(size = 1.4) +
  geom_vline(xintercept = 7, colour = "firebrick", linewidth = 0.5) +
  labs(x = "Circadian acrophase (h after the dose)",
       y = "AUC0-24h relative to daily dose / 23.0 L/h",
       title = "Sensitivity of steady-state AUC to the circadian phase",
       caption = paste("Red line: acrophase 7 h as printed in equation 1.",
                       "Dashed / dotted: the ratio Table 3 requires for 400 mg QD / 300 mg BID."))

summ <- tibble::tibble(
  Regimen = c("400 mg QD", "300 mg BID"),
  `Table 3 requires` = c(19034 / (400 / 23 * 1000), 29669 / (600 / 23 * 1000)),
  `At acrophase 7 (equation 1)` = c(phase$`400 mg QD`[phase$acrophase == 7],
                                    phase$`300 mg BID`[phase$acrophase == 7]),
  `Best over any acrophase` = c(max(phase$`400 mg QD`), max(phase$`300 mg BID`))
)
knitr::kable(summ, digits = 4,
             caption = "AUC0-24h relative to daily dose / mesor: what Table 3 requires versus what equation 1 can produce.")
```

The once-daily gap **is** a phase effect: Table 3 requires a ratio of 1.094 and
the model reaches 1.098 at an acrophase near 19 h -- a 12 h shift from the 7 h
printed in equation 1. The twice-daily gap is not: Table 3 requires 1.137 and
**no** acrophase gets the twice-daily regimen above about 1.016, because the
12 h dosing interval averages the 24 h rhythm out.

Table 3 therefore cannot be reproduced from equations 1-7 as printed, under any
phase convention. The model file encodes equation 1 exactly as published --
acrophase 7 h, which the paper's own 16:00 statement corroborates -- and this
disagreement is recorded rather than tuned away. Note that every quantity the
paper states in prose (both half-lives, both clearances, the Vc/F range, the
total V/F, tmax, and the *relative* comparison between regimens) is reproduced;
it is only the absolute level of the Table 3 Monte Carlo medians that differs.

```{r relative-claims}
med <- ss_res |> filter(PPTESTCD %in% c("cmax", "auclast", "clast.obs")) |>
  select(regimen, PPTESTCD, median) |>
  pivot_wider(names_from = PPTESTCD, values_from = median)

g <- function(r, p) med[[p]][med$regimen == r]
rel <- tibble::tibble(
  Quantity = c("AUC0-24h", "Cmax", "Ctrough"),
  `Paper: 400 mg QD vs 300 mg BID (%)` = c(-36, -18, -38),
  `Model (%)` = c(100 * (g("400 mg QD", "auclast")   / g("300 mg BID", "auclast")   - 1),
                  100 * (g("400 mg QD", "cmax")      / g("300 mg BID", "cmax")      - 1),
                  100 * (g("400 mg QD", "clast.obs") / g("300 mg BID", "clast.obs") - 1))
)
knitr::kable(rel, digits = 1,
             caption = "Kang 2025 Results section 2.4 / Discussion: 400 mg once daily versus the approved 300 mg twice daily.")

# Cohort-derived. Gate the direction and the magnitude band the paper claims,
# not the exact percentage. Realised -39 / -22 / -48 locally.
stopifnot(all(rel$`Model (%)` < -10), all(rel$`Model (%)` > -60))
```

## Single-dose healthy volunteers

```{r pknca-hv}
hv <- sim |> filter(DIS_CML == 0) |> transmute(id, regimen, time, Cc)

hv <- bind_rows(
  hv,
  hv |> distinct(id, regimen) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, regimen, time, .keep_all = TRUE) |>
  arrange(id, time)

hv_dose <- events |>
  filter(DIS_CML == 0, evid == 1) |>
  transmute(id, regimen, time, amt)

hv_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(as.data.frame(hv), Cc ~ time | regimen + id,
                   concu = "ng/mL", timeu = "h"),
  PKNCA::PKNCAdose(as.data.frame(hv_dose), amt ~ time | regimen + id,
                   doseu = "mg"),
  intervals = data.frame(start = 0, end = 48, cmax = TRUE, tmax = TRUE,
                         auclast = TRUE, half.life = TRUE)
))

hv_tab <- as.data.frame(hv_res$result) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "half.life")) |>
  group_by(PPTESTCD) |>
  summarise(Median = median(PPORRES, na.rm = TRUE),
            Q1 = quantile(PPORRES, 0.25, na.rm = TRUE),
            Q3 = quantile(PPORRES, 0.75, na.rm = TRUE), .groups = "drop") |>
  mutate(Parameter = nlmixr2lib::ncaParamLabel(PPTESTCD)) |>
  select(Parameter, Median, Q1, Q3)

knitr::kable(hv_tab, digits = 1,
             caption = "Single 400 mg oral dose in 150 virtual healthy volunteers, 0-48 h.")
```

The NCA half-life over a 48 h window is shorter than the model's own terminal
half-life of 18.2 h, because 48 h of sampling after an oral dose with a
1.88 h transit-absorption delay does not fully resolve the terminal phase --
this is a property of the sampling schedule (Kang 2025 Methods section 4.1
samples to 48 h), not of the model. The paper makes the same point in reverse
in its Discussion, noting that its model-derived CL/F of 37.9 L/h is slower than
the 51.0 L/h obtained by NCA on the same data.

```{r hv-gate}
hv_cmax <- hv_tab$Median[hv_tab$Parameter == nlmixr2lib::ncaParamLabel("cmax")]
hv_tmax <- hv_tab$Median[hv_tab$Parameter == nlmixr2lib::ncaParamLabel("tmax")]

# Cohort-derived; assert the shape of the profile, not a published number
# (Kang 2025 reports no NCA table for the healthy-volunteer arm).
stopifnot(hv_tmax >= 1.5, hv_tmax <= 6, hv_cmax > 500, hv_cmax < 5000)
```

# Assumptions and deviations

* **Table 3 is not reproducible from equations 1-7 as printed.** Simulated
  steady-state exposures sit about 12% (once daily) and 9% (twice daily) below
  the published Monte Carlo medians on AUC0-24h, with Cmax about 10% low and
  Ctrough within 6% for the once-daily arms.
  Table 3 is internally perfectly dose-proportional and implies an effective
  CL/F of 21.0 L/h (once daily) and 20.2 L/h (twice daily) against the 23.0 L/h
  mesor of equation 1. The once-daily gap is reproduced by a 12 h shift of the
  circadian acrophase, but no acrophase reproduces the twice-daily gap. The
  model encodes equation 1 exactly as published; the disagreement is recorded,
  not tuned. Everything the paper states in prose -- both clearances, both
  terminal half-lives, the Vc/F range, total V/F, tmax, and the relative
  comparison between regimens -- is reproduced.
* **The omega column is read as a standard deviation.** Kang 2025 Table 2 does
  not label the scale of its IIV column, but Results section 2.3 states that the
  MTT and N variances were fixed at 0.1 and 0.2 and Table 2 prints 0.316 and
  0.447, which are their square roots. Every entry is squared in the model file.
* **The IOV eta is shared between Vc/F and Vp/F.** Equations 2 and 4 both
  multiply by the same symbol `e^(eta_IOV)` with no distinguishing subscript,
  and Table 2 reports a single `omega IOV` row, so one occasion-level random
  effect drives both volumes. The alternative reading -- two independent etas
  with a common variance -- is not excluded by the text but is not what the
  equations print.
* **The circadian acrophase is a structural constant.** The 7 h phase shift
  appears only inside equation 1; Table 2 neither tabulates it nor reports an
  RSE for it, so it is encoded as `fixed(7)`. `TIME` is read as hours since the
  first dose (not clock time), which is the only reading consistent with the
  Discussion's statement that CL/F peaks at 16:00. Simulations that dose at a
  clock time other than ~09:00 should shift `acrophase_cl` accordingly.
* **Transit absorption uses the Savic 2007 parameterisation**, `ktr = (N + 1)/MTT`.
  Kang 2025 estimates MTT and N and states that a "transit compartment model
  with first-order absorption" was used (Methods section 4.3), but does not
  print the transit equations. The `(N + 1)` convention -- in which the mean
  transit time spans the N transit compartments plus the depot they feed -- is
  the standard form for this model class and the one the separately estimated
  `ka` implies. The chain is implemented in the analytical gamma-kernel form so
  that the non-integer N = 6.58 is handled smoothly.
* **The analytical transit kernel tracks the most recent dose only.** `podo()`
  and `tad()` restart at each dose, so superposition of overlapping absorption
  from consecutive doses is not represented. With MTT = 1.88 h and
  ktr = 4.03 1/h the chain is essentially exhausted within about 5 h, well
  inside the shortest (12 h) dosing interval used here, so the approximation is
  immaterial for these regimens.
* **No IIV on Vc/F, Q/F, Vp/F or ka.** Kang 2025 Results section 2.3 fixed those
  variances at exactly zero, so no eta is carried for them. A zero-variance eta
  would make the OMEGA matrix singular and break rxSolve's Cholesky sampler.
* **Age distribution.** Kang 2025 Table 1 reports only medians and ranges, so
  the virtual cohorts draw age from a truncated normal matched to the per-arm
  median and range (CML-CP median 32, range 21-72; healthy volunteers median 29,
  range 20-51). The shape between those points is an assumption. Age enters only
  Vc/F, which does not affect steady-state exposure.
* **Weight, sex, ethnicity, hepatic and renal function are not in the model.**
  They were screened (Methods section 4.3) but not retained, so no covariate
  columns are carried for them.
* **The disease effect is confounded.** Kang 2025's own Discussion (limitations)
  notes that all healthy volunteers were Korean males and all patients were
  Chinese of both sexes, so the 0.646 coefficient may partly reflect interethnic
  or sex-related variability rather than disease status alone. The coefficient
  is encoded as published; users should not transport it to other populations
  without that caveat.
* **The linear age effect must not be extrapolated.** `1 - 0.0129 * (AGE - 31)`
  reaches zero at age 108.5 and turns negative beyond it. It is supported only
  over the observed 20-72 year range.
* **All values are from the paper's text and tables.** No supplement was
  published for this article and no parameter was digitised from a figure or
  obtained by correspondence.
