---
title: "Octreotide LAR (Kapralos 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Octreotide LAR (Kapralos 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Kapralos_2021_octreotide"))$reference`
* Description: `r rxode2::rxode(readModelDb("Kapralos_2021_octreotide"))$description`
* Article: <https://doi.org/10.3390/pharmaceutics13101578>
* Supplement (base-model Table S1 and goodness-of-fit figures): <https://www.mdpi.com/article/10.3390/pharmaceutics13101578/s1>

Octreotide LAR (Sandostatin LAR Depot) releases octreotide from PLGA
microspheres over about two months. Kapralos and Dokoumetzidis describe the
multi-peak release with an empirical input model: a small initial burst plus
three parallel delayed releases, each a Savic transit-compartment density,
feeding a first-order absorption compartment and a one-compartment
disposition model (Figure 1 of the paper). Before the population fit, a
shape-respecting k-means clustering of the individual profiles split the
subjects into a typical multi-phase group (cluster 1) and an early
extended-release group (cluster 2). The cluster is carried as the binary
covariate `MIX_EARLY_REL` (0 = cluster 1, 1 = cluster 2).

# Population

118 healthy adult male Caucasian volunteers from the reference arm of a
phase 1 single-dose bioequivalence study received one 30 mg deep
intramuscular injection of Sandostatin LAR Depot under fasting conditions.
Serum octreotide was sampled pre-dose and at 36 times from 0.5 h to 2088 h
(3936 observations). Median (Q1-Q3) age was 28 (23-37) years, weight 75
(66-86) kg, height 175 (170-178) cm and BMI 24.75 (22.4-27.7) kg/m^2
(Table 1). The clustering assigned 103 subjects to cluster 1 and 15 (12.7%)
to cluster 2 (Table 1).

The same information is available programmatically via
`readModelDb("Kapralos_2021_octreotide")()$population`.

# Source trace

Every `ini()` value carries an in-file comment pointing to its source in
`inst/modeldb/specificDrugs/Kapralos_2021_octreotide.R`. The table collects
them.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` | log(0.27) 1/h | Table 2, ka |
| `lcl` | log(32.7) L/h | Table 2, CL |
| `lvc` | log(15.3) L | Table 2, V |
| `e_mix_early_rel_cl` | -8.61 L/h | Table 2, CL cluster effect; Equation 7 (additive) |
| `logitfburst` | -5.18 | Table 2, YF1 |
| `logitfdel1` | -3.36 | Table 2, YF2 |
| `logitfdel2` | -1.54 | Table 2, YF3 |
| `e_mix_early_rel_logitfdel1` | 3.06 | Table 2, YF2 cluster effect |
| `e_mix_early_rel_logitfdel2` | -0.523 | Table 2, YF3 cluster effect |
| `mtt1max` | 300 h (fixed) | Section 2.2.4, Equation 4 |
| `logitmtt1` | -0.421 | Table 2, YMTT1 |
| `ldmtt2` | log(181) h | Table 2, MTT2; Equation 3 increment |
| `ldmtt3` | log(506) h | Table 2, MTT3; Equation 3 increment |
| `lnn1`, `lnn2`, `lnn3` | log(3.42), log(17.9), log(5.08) | Table 2, N1-N3 |
| IIV variances | (CV/100)^2 | Table 2, IIV rows; CV definition in Section 2.2.4 |
| `propSd` | 0.143 | Table 2, proportional residual error |
| `addSd` | 28.4 pg/mL | Table 2, additive residual error |
| Release fractions `fburst`-`fdel3` | n/a | Equation 5 and Figure 1 |
| `mtt1` logit constraint | n/a | Equation 4 |
| `mtt2 = mtt1 + dmtt2`, `mtt3 = mtt2 + dmtt3` | n/a | Equation 3 |
| `ktr = (n + 1) / MTT` | n/a | Figure 1 |
| Transit density with Stirling n! | n/a | Equation 2 |
| `d/dt(depot)` | n/a | Equation 1; burst bolus from Figure 1 |
| Cluster effects additive on the typical value | n/a | Equation 7 |

# Structural checks

## Typical release fractions and transit times

```{r typical-params}
mod <- readModelDb("Kapralos_2021_octreotide")
mod_typ <- rxode2::zeroRe(mod)

tgrid <- sort(unique(c(seq(0, 48, by = 0.1), seq(49, 6000, by = 1))))
make_typ <- function(mix, id) {
  dplyr::bind_rows(
    data.frame(id = id, time = 0, evid = 1, amt = 30, cmt = "depot"),
    data.frame(id = id, time = tgrid, evid = 0, amt = 0, cmt = "central")
  ) |>
    dplyr::mutate(MIX_EARLY_REL = mix)
}
ev_typ <- dplyr::bind_rows(make_typ(0, 1L), make_typ(1, 2L))
sim_typ <- rxode2::rxSolve(mod_typ, events = ev_typ, returnType = "data.frame") |>
  dplyr::mutate(cluster = ifelse(id == 1, "Cluster 1", "Cluster 2"))

typ_par <- sim_typ |>
  dplyr::group_by(cluster) |>
  dplyr::slice(1) |>
  dplyr::ungroup() |>
  dplyr::select(cluster, cl, fburst, fdel1, fdel2, fdel3, mtt1, mtt2, mtt3, nn1, nn2, nn3)
knitr::kable(typ_par, digits = 4, caption = "Typical-subject parameters by cluster.")
```

For cluster 1 the burst carries 0.45% of the dose, the first delayed release
2.8%, the second 17.1% and the third 79.7%. The paper's statement that the
initial burst accounts for less than 1/100 of the total exposure (Discussion)
matches `fburst`.

## Sequential mean transit times (Equation 3)

Table 2 lists "MTT2" = 181 h and "MTT3" = 506 h. Equation 3 defines
`MTT_j = MTT_(j-1) + theta_j * exp(eta_j)`, so the model treats the two
Table 2 values as the increments `theta_2` and `theta_3`, which gives
MTT1 = 119 h, MTT2 = 300 h and MTT3 = 806 h. The published profiles support
this reading. The mode of a Savic input is at `n / ktr = n * MTT / (n + 1)`,
so the sharp second delayed release (N2 = 17.9) peaks at about 284 h under
the increment reading. If 181 h were the absolute MTT2, it would peak at
about 171 h. Figure 2 (cluster-1 mean profile) and Figure 5 (VPC median) both
show the sharp second peak at about 300 h and a broad plateau lasting past
1000 h. The broad plateau is what the third release produces with a mean of
806 h, not 506 h.

```{r mtt-check}
typ1 <- sim_typ |> dplyr::filter(cluster == "Cluster 1", time > 48)
y <- typ1$Cc
local_max <- typ1$time[which(diff(sign(diff(y))) == -2) + 1]
local_max
# Deterministic typical-value gate: the second delayed peak must sit near the
# ~300 h peak of Figures 2 and 5, not at the ~171 h an absolute-MTT reading
# would put it.
stopifnot(any(local_max > 250 & local_max < 350))
```

## Mass balance

The closed-form input uses Stirling's approximation of `n!` (Equation 2), so
each delayed release delivers `Gamma(n + 1) / Stirling(n)` of its fraction,
slightly more than one (about `1 + 1/(12 n)`). For every subject the model must
therefore satisfy
`CL * AUC(0-inf) = Dose * (fburst + sum_j f_j * Gamma(n_j + 1) / Stirling(n_j))`.
Both sides use the same drawn parameters, so the check is tight. A mutation
control that drops the Stirling factor must fail, which shows the gate can go
red.

```{r mass-balance}
stir_factor <- function(n) exp(lgamma(n + 1) - (0.5 * log(2 * pi) + (n + 0.5) * log(n) - n))
mb <- sim_typ |>
  dplyr::group_by(cluster) |>
  dplyr::summarise(
    auc = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2) / 1e6, # mg*h/L
    cl = cl[1],
    expected = 30 * (fburst[1] + fdel1[1] * stir_factor(nn1[1]) +
      fdel2[1] * stir_factor(nn2[1]) + fdel3[1] * stir_factor(nn3[1])),
    naive = 30,
    .groups = "drop"
  ) |>
  dplyr::mutate(ratio = cl * auc / expected, ratio_naive = cl * auc / naive)
knitr::kable(mb, digits = 5)
stopifnot(
  all(abs(mb$ratio - 1) < 1e-3),
  # mutation control: ignoring the Stirling factor is detectably wrong
  all(abs(mb$ratio_naive - 1) > 5e-3)
)
```

The ratio of the typical AUCs is set by the cluster effect on CL,
32.7 / (32.7 - 8.61) = 1.357, times the small difference in the delivered
amount (the two clusters split the dose differently across releases with
different Stirling factors). The observed cluster AUC ratio is
1295.2 / 944.0 = 1.372 (Table 1).

```{r auc-ratio}
typ_auc_ratio <- mb$auc[mb$cluster == "Cluster 2"] / mb$auc[mb$cluster == "Cluster 1"]
expected_ratio <- (32.7 / (32.7 - 8.61)) *
  mb$expected[mb$cluster == "Cluster 2"] / mb$expected[mb$cluster == "Cluster 1"]
c(simulated = typ_auc_ratio, expected = expected_ratio)
stopifnot(abs(typ_auc_ratio / expected_ratio - 1) < 1e-3)
```

# Typical-value profiles

```{r figure-typical, fig.width = 7, fig.height = 4}
sim_typ |>
  dplyr::filter(time <= 2100) |>
  ggplot(aes(time, Cc, colour = cluster)) +
  geom_line() +
  labs(
    x = "Time after dose (h)", y = "Octreotide (pg/mL)", colour = NULL,
    title = "Typical-subject profiles after 30 mg octreotide LAR",
    caption = "Compare with the population (PRED) lines of Figure 3 and the mean profiles of Figure 2 of Kapralos 2021."
  )
```

The input rates of the four release processes for a typical cluster-1
subject show how the processes combine (Figure 1 of the paper).

```{r figure-inputs, fig.width = 7, fig.height = 4}
sim_typ |>
  dplyr::filter(cluster == "Cluster 1", time <= 2100) |>
  dplyr::transmute(
    time,
    `Delayed release 1` = 30 * fdel1 * rin1,
    `Delayed release 2` = 30 * fdel2 * rin2,
    `Delayed release 3` = 30 * fdel3 * rin3
  ) |>
  tidyr::pivot_longer(-time, names_to = "process", values_to = "rate") |>
  ggplot(aes(time, rate * 1000, colour = process)) +
  geom_line() +
  labs(
    x = "Time after dose (h)", y = "Input rate (ug/h)", colour = NULL,
    title = "Delayed-release input rates, typical cluster-1 subject",
    caption = "The burst (0.45% of the dose) enters as a bolus at time 0 and is not shown."
  )
```

# Virtual cohort and VPC

The paper's VPC is stratified by cluster (Figure 5). The simulation uses 200
virtual subjects per cluster on the study's sampling schedule. Body size
and age were not retained as covariates, so the cohort needs no demographic
columns.

```{r cohort}
rxode2::rxSetSeed(20211)
samp_times <- c(
  0, 0.5, 1, 1.5, 2, 3, 4, 6, 10, 24, 48, 72, 96, 144, 192, 240, 288, 336,
  384, 432, 480, 528, 576, 624, 672, 720, 768, 816, 864, 912, 1008, 1176,
  1344, 1512, 1680, 1848, 2088
)
make_cohort <- function(n, mix, id_offset = 0L) {
  ids <- id_offset + seq_len(n)
  dplyr::bind_rows(
    data.frame(id = ids, time = 0, evid = 1, amt = 30, cmt = "depot"),
    expand.grid(id = ids, time = samp_times) |>
      dplyr::mutate(evid = 0, amt = 0, cmt = "central")
  ) |>
    dplyr::mutate(
      MIX_EARLY_REL = mix,
      cluster = ifelse(mix == 1, "Cluster 2", "Cluster 1")
    ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}
events <- dplyr::bind_rows(
  make_cohort(200, 0, id_offset = 0L),
  make_cohort(200, 1, id_offset = 200L)
)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

```{r simulate}
sim <- rxode2::rxSolve(mod, events = events, keep = "cluster", returnType = "data.frame")
```

```{r figure-5, fig.width = 7, fig.height = 4}
sim |>
  dplyr::group_by(cluster, time) |>
  dplyr::summarise(
    Q05 = quantile(sim, 0.05, na.rm = TRUE),
    Q50 = quantile(sim, 0.50, na.rm = TRUE),
    Q95 = quantile(sim, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~cluster) +
  labs(
    x = "Time (h)", y = "Octreotide (pg/mL)",
    title = "Simulated 5th, 50th and 95th percentiles by cluster",
    caption = "Replicates Figure 5 of Kapralos 2021 (VPC of the final model stratified on cluster)."
  )
```

# PKNCA validation

Table 1 reports arithmetic means of the observed AUC(0-t) and Cmax by
cluster. PKNCA runs on the simulated concentrations including residual error
(the `sim` column), because the observed values carry assay noise. Values
below the assay's lower calibration limit (8.835 pg/mL) are set to zero before
NCA. The paper's Figure 6 partial AUCs (0-24 h, 0-28 days and 28-56 days) are
computed as well.

```{r pknca}
lloq <- 8.835
sim_nca <- sim |>
  dplyr::filter(!is.na(sim)) |>
  dplyr::transmute(id, time, cluster, conc = ifelse(sim < lloq, 0, sim))
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, cluster) |> dplyr::mutate(time = 0, conc = 0)
) |>
  dplyr::distinct(id, cluster, time, .keep_all = TRUE) |>
  dplyr::arrange(id, cluster, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, conc ~ time | cluster + id)
dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, cluster)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | cluster + id)

intervals <- data.frame(
  start = c(0, 0, 0, 672),
  end = c(2088, 24, 672, 1344),
  cmax = c(TRUE, FALSE, FALSE, FALSE),
  auclast = c(TRUE, FALSE, FALSE, FALSE),
  aucint.last = c(FALSE, TRUE, TRUE, TRUE)
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_long <- as.data.frame(nca_res$result)
```

## Comparison against published NCA

```{r nca-compare}
# Table 1 reports arithmetic means; aggregate the simulated values the same way.
sim_means <- nca_long |>
  dplyr::filter(PPTESTCD %in% c("cmax", "auclast"), start == 0, end == 2088) |>
  dplyr::group_by(cluster, PPTESTCD) |>
  dplyr::summarise(PPORRES = mean(PPORRES, na.rm = TRUE), .groups = "drop")
# Whole cohort: weight the cluster means by the observed 103 / 15 split.
sim_means <- dplyr::bind_rows(
  sim_means,
  sim_means |>
    dplyr::group_by(PPTESTCD) |>
    dplyr::summarise(
      PPORRES = sum(PPORRES * ifelse(cluster == "Cluster 1", 103, 15)) / 118,
      .groups = "drop"
    ) |>
    dplyr::mutate(cluster = "All subjects")
)
published <- tibble::tribble(
  ~cluster,       ~cmax,  ~auclast,
  "Cluster 1",    1433.3, 944.0e3,
  "Cluster 2",    5034.8, 1295.2e3,
  "All subjects", 1891.1, 988.7e3
)
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_means,
  reference = published,
  by = "cluster",
  units = c(cmax = "pg/mL", auclast = "pg*h/mL"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = "Simulated vs. published (Table 1) arithmetic-mean NCA. * differs from reference by >20%."
)
```

The simulated means reproduce Table 1 closely in both clusters and in the
whole cohort; no row is flagged.

```{r nca-gate}
chk <- sim_means |>
  dplyr::inner_join(
    tidyr::pivot_longer(published, -cluster, names_to = "PPTESTCD", values_to = "ref"),
    by = c("cluster", "PPTESTCD")
  ) |>
  dplyr::mutate(pct_diff = 100 * (PPORRES / ref - 1))
chk
# Rendered values were within 1.1% (AUC) and 3.5% (Cmax) of Table 1. The
# sampling standard error of a 200-subject mean is about 2% for AUC and 2.5-4%
# for Cmax (heavy upper tail from the YF2 IIV), so the envelopes below leave
# room for a different cohort draw while still catching a mis-scaled unit or
# a wrong cluster effect, either of which moves the means by tens of percent.
stopifnot(
  all(abs(chk$pct_diff[chk$PPTESTCD == "auclast"]) < 10),
  all(abs(chk$pct_diff[chk$PPTESTCD == "cmax"]) < 25)
)
```

The bioequivalence metrics of Figure 6 are summarised below as simulated
10th, 50th and 90th percentiles. The paper shows these only graphically, so
they are not compared numerically.

```{r be-metrics}
nca_long |>
  dplyr::mutate(metric = dplyr::case_when(
    PPTESTCD == "cmax" ~ "Cmax (pg/mL)",
    PPTESTCD == "auclast" ~ "AUC(0-t) (pg*h/mL)",
    end == 24 ~ "AUC(0-24 h) (pg*h/mL)",
    end == 672 ~ "AUC(0-28 d) (pg*h/mL)",
    end == 1344 ~ "AUC(28-56 d) (pg*h/mL)"
  )) |>
  dplyr::filter(!is.na(metric), PPTESTCD %in% c("cmax", "auclast", "aucint.last")) |>
  dplyr::group_by(cluster, metric) |>
  dplyr::summarise(
    P10 = signif(quantile(PPORRES, 0.1, na.rm = TRUE), 3),
    P50 = signif(quantile(PPORRES, 0.5, na.rm = TRUE), 3),
    P90 = signif(quantile(PPORRES, 0.9, na.rm = TRUE), 3),
    .groups = "drop"
  ) |>
  dplyr::rename("Cluster" = cluster, "Metric" = metric) |>
  knitr::kable(caption = "Simulated bioequivalence metrics (compare Figure 6 of Kapralos 2021).")
```

# Assumptions and deviations

- **MTT2 and MTT3 are increments.** Table 2 labels the two estimates "MTT2"
  and "MTT3", but Equation 3 defines each later mean transit time as the
  previous one plus `theta_j * exp(eta_j)`. The model uses the Table 2 values
  as those increments. The typical profile then puts the sharp second delayed
  peak at about 300 h and a plateau past 1000 h, as in Figures 2 and 5. Under
  the alternative reading the second peak would be at about 171 h and the
  third release would be exhausted by about 1000 h. See the structural checks
  above.
- **IIV scale.** Section 2.2.4 defines the reported IIV as
  `CV(%) = sqrt(omega^2) * 100`, so each variance is `(CV/100)^2`, including
  the logit-normal fraction and MTT1 parameters.
- **IIV covariances not reported.** Section 3.2 states that the covariances
  between the fraction and the mean-transit-time etas were significant and
  kept. Neither Table 2 nor the supplementary Table S1 prints them, so the
  model carries a diagonal omega. Simulated between-subject spread of the
  release shape may therefore differ from the published VPC.
- **Residual error scale.** Table 2 lists the proportional (0.143) and
  additive (28.4 pg/mL) residual errors among the fixed effects. The model
  treats them as standard deviations of a combined error model. The paper does
  not say how the two components were combined, so nlmixr2's default
  `add() + prop()` is used.
- **Stirling approximation kept.** Equation 2 replaces `n!` in the Savic
  density with Stirling's approximation. The model keeps this form because the
  parameters were estimated with it. As a result the total delivered amount
  slightly exceeds the dose (by about 1.5% for a typical subject); the
  mass-balance check above accounts for this exactly.
- **No IIV on ka.** Section 3.2 says IIV was estimated for every parameter
  except ka. The base model in Table S1 shows ka IIV as `0 FIXED`.
- **Single-dose model.** The delayed-release inputs read the most recent dose
  amount and time after dose (`podo()` / `tad()`). With repeated monthly
  injections, a new dose would cut off the still-running release of the
  previous one (the release lasts about 2000 h). The paper studied a single
  dose only. Multiple-dose use would need superposition, which this model does
  not provide.
- **Cluster covariate.** The cluster assignment comes from a pre-fit
  clustering of each subject's observed profile. It cannot be measured before
  dosing. For population simulation, draw `MIX_EARLY_REL ~ Bernoulli(15/118)`.
- **NCA inputs.** The virtual cohort is 200 subjects per cluster. The
  whole-cohort NCA means weight the two clusters by the observed 103 / 15
  split. Simulated concentrations below 8.835 pg/mL (the lowest calibration
  standard) were set to zero for NCA. The paper does not describe its
  below-limit handling.
- **Errata.** Crossref lists no correction or update to this article as of
  2026-09-29.
