---
title: "Acute stroke disease progression on the NIHSS and Barthel Index (Karlsson 2010)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Acute stroke disease progression on the NIHSS and Barthel Index (Karlsson 2010)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Karlsson 2010 develops two independent disease-progression models -- one per
clinical assessment scale -- from the same placebo cohort. They are packaged as
two model files, and this single vignette walks both.

* NIHSS model: `r rxode2::rxode(readModelDb("Karlsson_2010_acute_stroke_nihss"))$reference`
* Barthel Index model: same publication, Supplement A BI table.
* Article: <https://doi.org/10.1208/s12248-010-9230-0> (PMC free full text: PMC2977005)
* Supplement A (parameter estimates): <https://static-content.springer.com/esm/art%3A10.1208%2Fs12248-010-9230-0/MediaObjects/12248_2010_9230_MOESM1_ESM.docx>
* Supplement B (NONMEM control streams): <https://static-content.springer.com/esm/art%3A10.1208%2Fs12248-010-9230-0/MediaObjects/12248_2010_9230_MOESM2_ESM.docx>

```{r load-models}
nihss <- readModelDb("Karlsson_2010_acute_stroke_nihss")
bi    <- readModelDb("Karlsson_2010_acute_stroke_bi")

ui_nihss <- rxode2::rxode(nihss)
ui_bi    <- rxode2::rxode(bi)
```

There is **no drug, no dose and no ODE** anywhere in this paper. The source is a
NONMEM `$PRED` model fitted to the placebo arm, so both packaged models are
purely algebraic: they map a set of Markovian predictors supplied as data
columns onto five simultaneous outputs.

# Population

`r rxode2::rxode(readModelDb("Karlsson_2010_acute_stroke_nihss"))$population$n_subjects` acute
stroke patients from the placebo arm of the double-blind, multinational,
multicentre CLASS-I trial of clomethiazole (Karlsson 2010 reference 32 = Lyden
et al., *Stroke* 2002;33:122-128). Patients were eligible if stroke onset was
within 12 h of treatment. Mean age 71.7 years (range 26-90); mean baseline
NIHSS 16.8 (range 4-31) (Karlsson 2010 Methods, *Patients and Data*).

The NIHSS was assessed at admission and on days 7, 30 and 90. The Barthel Index
was assessed on days 7, 30 and 90 plus a telephone assessment at day 60; it was
**not** assessed at admission "for practical and ethical reasons", and the
authors tested static and randomly-imputed baselines between 0 and 20 before
rejecting them for insufficient predictive power. The BI is therefore modelled
from day 7 onward.

The population PK model from this same trial is packaged separately as
`Zingmark_2003_clomethiazole`; Karlsson 2010 is its placebo-arm
disease-progression companion.

Full metadata is available programmatically:

```{r population}
str(rxode2::rxode(readModelDb("Karlsson_2010_acute_stroke_nihss"))$population)
```

# Model structure

Both scales share one structure. At every observation occasion the model asks a
**hierarchy** of questions (Karlsson 2010 Figures 1 and 2), then -- if the
subject neither fully recovered nor dropped out -- predicts *how much* the score
moved:

1. `p_improve` -- did the score improve, or decline? (An unchanged score counts
   as a decline, per Karlsson 2010 Data Preparation.)
2. If it improved: `p_notmax` -- did it stop short of the fully-recovered score,
   or reach it?
3. If it declined: `p_dropout` -- did the subject drop out?
4. `imz` / `dez` -- the *relative* magnitude of the improvement or decline, on
   the logit scale, back-transformed onto the raw scale by the authors'
   reconstruction rules.

The hierarchy is not a modelling choice made here; it is visible in the source
dataset. In the Supplement B NIHSS excerpt, subject 1 at day 7 carries a
`TYPE 1` record with `DV = 0` (decline) and a `TYPE 3` record (dropout) but no
`TYPE 2`; at day 89 it carries `TYPE 1` with `DV = 1` (improvement) and a
`TYPE 2` record (maximum) but no `TYPE 3`. Recovery is asked only after an
improvement, dropout only after a decline.

The reconstruction rules differ because the two scales run in opposite
directions -- **lower NIHSS is better, higher BI is better**:

| | NIHSS (0-42, lower better) | Barthel Index (0-100, higher better) |
|---|---|---|
| Improvement | `nihss_improve = (1 - imz) * (SCORE_NIHSS_PREV + 1)` | `bi_improve = SCORE_BI_PREV + imz * (100 - SCORE_BI_PREV)` |
| Decline | `nihss_decline = dez * (42 - SCORE_NIHSS_PREV) + SCORE_NIHSS_PREV` | `bi_decline = (1 - dez) * (SCORE_BI_PREV + 1)` |
| Fully recovered | score set to 0 | score set to 100 |

Dropout is **informative** in both models: its logit enters proportionally on
the model-predicted *declined* score, `logitp_dropout * (1 + slope_score_p_dropout * <declined score>)`
(Karlsson 2010 Table II footnote b). A patient predicted to decline further is
more likely to leave the study, and that information is used rather than
discarded.

The two models differ in where they place their time dependence. The NIHSS model
carries three day break points and four IIV terms; the BI model carries none of
those but does carry two distinct clocks:

```{r structure-diff, echo = FALSE}
tibble::tribble(
  ~Feature,                                       ~NIHSS,                                  ~`Barthel Index`,
  "Break point on relative improvement",          "day 14",                                "none",
  "Break point on p(not maximum)",                "day 45",                                "none",
  "Break points on p(dropout)",                   "day 14 and day 45 (3 intervals)",       "none",
  "IIV terms",                                    "4 (improvement and decline, each split at day 14)", "1 (decline only)",
  "Time since previous observation (T_PREVOBS)",  "on p(not maximum)",                     "on relative decline",
  "Time since stroke (rxode2 `time`)",            "break points only",                     "on p(not maximum)"
) |>
  knitr::kable(caption = "Where the two scale models differ (Karlsson 2010 Table II and the Supplement B control streams).")
```

# Source trace

Every `ini()` entry carries an in-file comment naming its Supplement A row. The
table below collects them.

**Supplement A holds the final estimates; Supplement B's `$THETA` / `$OMEGA`
records are NONMEM *initial* estimates and are not used here.** The two disagree
widely -- on the BI, 11 of 15 thetas and the omega differ (1.25 vs 2.08). The
paper's own Results text adjudicates: it states that the NIHSS probability of
improvement "decreased 0.05 units with 10 years, between ages 64 and 74".
Supplement A reproduces that (0.6506 to 0.5978, a difference of 0.0527);
Supplement B's initial values give 0.0411. Supplement B is used for **model
structure only**.

```{r source-trace, echo = FALSE}
trace_tbl <- function(ui, tbl_label) {
  df <- ui$iniDf
  df <- df[is.na(df$neta1) | df$neta1 == df$neta2, ]
  tibble::tibble(
    Parameter = df$name,
    Value     = signif(df$est, 4),
    Kind      = ifelse(is.na(df$neta1), ifelse(df$fix, "THETA (fixed)", "THETA"), "OMEGA"),
    Source    = tbl_label
  )
}

dplyr::bind_rows(
  trace_tbl(ui_nihss, "Karlsson 2010 Supplement A, NIHSS table"),
  trace_tbl(ui_bi,    "Karlsson 2010 Supplement A, BI table")
) |>
  dplyr::rename("Source location" = Source) |>
  knitr::kable(caption = "Source trace: every packaged parameter and its Supplement A row.")
```

Two structural elements come from Supplement B rather than Supplement A, because
they are not parameter values:

| Element | Source location |
|---|---|
| Five-submodel hierarchy, `P = exp(lambda)/(1 + exp(lambda))` | Karlsson 2010 Figures 1-2; Methods, *Modeling Approach* |
| Reconstruction rules (`NIM2`, `NDE`, `BIM1`, `BDE2`) | Supplement B `$PRED`, "Transforming IMZ/DEZ to real NIH score" block; Karlsson 2010 Table I |
| Proportional informative-dropout form `THETA(3)*(1 + THETA(7)*NDE)` | Supplement B `$PRED` dropout block; Table II footnote b |
| Day break points (14, 45) and the eta-to-interval mapping | Supplement B `$PRED` `IF(DAY.GE.14)` / `IF(DAY.GT.45)` branches |
| Residual SD is `THETA(6)` itself | Supplement B `$SIGMA 1 FIX` with `RV = THETA(6)*EPS(1)` |

# Reproducing the effect sizes the paper prints

Karlsson 2010's Results section quotes four numeric effect sizes. These are the
sharpest available gate on the transcription: each is computed here **from the
packaged model** at typical values, not from a re-implementation of the
equations.

```{r effect-sizes}
solve_typical <- function(ui, ev) {
  as.data.frame(rxode2::rxSolve(rxode2::zeroRe(ui), ev, omega = NA))
}

## Claim 1 -- NIHSS: p(improvement) falls 0.05 between ages 64 and 74.
c1 <- solve_typical(ui_nihss, data.frame(
  id = 1:2, time = 30, evid = 0, amt = NA_real_, AGE = c(64, 74),
  SCORE_NIHSS = 16.8, SCORE_NIHSS_PREV = 16.8, T_PREVOBS = 23
))
claim1 <- c1$p_improve[1] - c1$p_improve[2]

## Claims 2 and 4 are both stated per unit of the model-PREDICTED declined
## score, which is a reconstruction of the previous score rather than an input.
## Invert the reconstruction through the model with uniroot() so the two
## evaluation points are exact, instead of stepping the previous score by the
## nominal amount (which would move the predicted score by less).
solve_nihss <- function(prev) solve_typical(ui_nihss, data.frame(
  id = seq_along(prev), time = 30, evid = 0, amt = NA_real_, AGE = 71.7,
  SCORE_NIHSS = 16.8, SCORE_NIHSS_PREV = prev, T_PREVOBS = 23
))
solve_bi <- function(prev) solve_typical(ui_bi, data.frame(
  id = seq_along(prev), time = 30, evid = 0, amt = NA_real_, AGE = 71.7,
  SCORE_BI_PREV = prev, SCORE_NIHSS = 16.8, T_PREVOBS = 23
))

invert_to <- function(f, column, target, interval) {
  uniroot(function(p) f(p)[[column]] - target, interval)$root
}

## Claim 2 -- NIHSS: p(dropout) rises ~0.015 per 1-point increase in the
## predicted declined score. The logistic is non-linear, so the increment
## depends on where it is evaluated; anchor it at the cohort mean baseline
## NIHSS of 16.8 (Karlsson 2010 Methods).
prev_n <- vapply(c(16.8, 17.8),
                 function(x) invert_to(solve_nihss, "nihss_decline", x, c(0, 42)),
                 numeric(1))
c2 <- solve_nihss(prev_n)
claim2 <- c2$p_dropout[2] - c2$p_dropout[1]

## Claim 3 -- BI: p(improvement) falls 0.07 between ages 64 and 74 at a
## previous BI of 50.
c3 <- solve_typical(ui_bi, data.frame(
  id = 1:2, time = 30, evid = 0, amt = NA_real_, AGE = c(64, 74),
  SCORE_BI_PREV = 50, SCORE_NIHSS = 16.8, T_PREVOBS = 23
))
claim3 <- c3$p_improve[1] - c3$p_improve[2]

## Claim 4 -- BI: p(dropout) falls 0.035 per 5-point increase in the
## predicted declined BI score.
prev_b <- vapply(c(20, 25),
                 function(x) invert_to(solve_bi, "bi_decline", x, c(0, 100)),
                 numeric(1))
c4 <- solve_bi(prev_b)
claim4 <- c4$p_dropout[2] - c4$p_dropout[1]
```

The two dropout claims are evaluated at previous scores of
`r sprintf("%.1f and %.1f", prev_n[1], prev_n[2])` (NIHSS) and
`r sprintf("%.1f and %.1f", prev_b[1], prev_b[2])` (Barthel Index), which are
the values the model maps onto predicted declined scores of 16.8/17.8 and
20/25 respectively.

```{r effect-size-table, echo = FALSE}
claims <- tibble::tribble(
  ~Claim, ~Published, ~Model, ~Deviation,
  "NIHSS: change in p(improvement), age 64 vs 74",                     0.05,   claim1, FALSE,
  "NIHSS: change in p(dropout) per +1 predicted score (score 16.8, day 30)", 0.015, claim2, FALSE,
  "BI: change in p(improvement), age 64 vs 74 at previous BI 50",       0.07,   claim3, TRUE,
  "BI: change in p(dropout) per +5 predicted BI points",               -0.035,  claim4, FALSE
) |>
  dplyr::mutate(
    Model      = round(Model, 4),
    Difference = round(Model - Published, 4)
  )

claims |>
  dplyr::rename(
    "Published value" = Published,
    "Model value"     = Model,
    "Known deviation" = Deviation
  ) |>
  knitr::kable(caption = "Effect sizes quoted in Karlsson 2010 Results, recomputed from the packaged models.")
```

Three of the four reproduce closely. The gate below is deterministic -- it runs
at typical values on a fixed covariate grid, so there is no cohort draw and the
tolerances can be tight.

```{r effect-size-gate}
stopifnot(
  # Claim 1: paper prints 0.05 to one significant figure; model gives 0.0527.
  abs(claim1 - 0.05) < 0.005,
  # Claim 2: paper says "approximately 0.015"; the model gives 0.0171 at the
  # cohort mean predicted score of 16.8. The increment is not constant -- the
  # logistic is steepest near p = 0.5 -- so the tolerance admits the range the
  # paper's unstated reference score could plausibly have been taken at.
  abs(claim2 - 0.015) < 0.005,
  # Claim 4: paper prints -0.035; model gives -0.0348.
  abs(claim4 - (-0.035)) < 0.002
)
```

**Known deviation, claim 3.** The BI age effect reproduces as
`r sprintf("%.4f", claim3)` against a published 0.07 -- close, but outside the
rounding of the printed value. It is not a cohort-draw artefact (the calculation
is deterministic) and it is not sensitive to the centring constant: recomputing
with age centred at the cohort mean of 71.7 rather than the source's 73 still
gives 0.081. Supplement B's *initial* estimates would give 0.0725, which is
nearer the printed 0.07 -- but adopting them would break claim 1 badly
(0.0411 against a published 0.05), and Supplement A is the table the paper
labels "parameter estimates ... standard errors and a log-likelihood profiling
produced 95% confidence interval". The reading taken here is that the paper's
Results prose rounded 0.0795 down rather than up. It is recorded rather than
tuned away.

# Virtual cohort

Original observed data are not publicly available. The cohort below matches the
published demographics: mean age 71.7 years truncated to the reported 26-90
range, and mean baseline NIHSS 16.8 truncated to the reported 4-31 range. The
paper reports means and ranges but no standard deviations, so the dispersions
are assumptions (see *Assumptions and deviations*).

```{r cohort}
# set.seed() seeds R's RNG. This vignette draws the between-subject etas, the
# residuals and the Bernoulli transition events in R rather than in rxode2, so
# set.seed() does govern the cohort here -- but it still governs it only for a
# given R version and RNG kind, so every assertion below is written to hold for
# any cohort the model can produce.
set.seed(20101)

N_SUBJ <- 200L  # per-arm cap; one arm here (placebo)

rtrunc_norm <- function(n, mean, sd, lo, hi) {
  pmin(hi, pmax(lo, rnorm(n, mean, sd)))
}

cohort <- tibble::tibble(
  id          = seq_len(N_SUBJ),
  AGE         = rtrunc_norm(N_SUBJ, 71.7, 12, 26, 90),
  SCORE_NIHSS = round(rtrunc_norm(N_SUBJ, 16.8, 5, 4, 31))
)

summary(cohort[, c("AGE", "SCORE_NIHSS")])
```

Between-subject etas are drawn once per subject from each model's own OMEGA and
then held fixed across that subject's visits. They are passed to `rxSolve()` as
data columns with `omega = NA`, so the packaged model -- not this vignette --
decides which eta applies at which visit (the NIHSS model switches all four at
day 14).

```{r draw-etas}
draw_etas <- function(ui, n) {
  om <- ui$omega
  # Both models have diagonal OMEGA; assert it so a future correlated block is
  # not silently sampled as independent.
  stopifnot(isTRUE(all.equal(unname(om), diag(diag(om), nrow = nrow(om)))))
  as.data.frame(vapply(
    colnames(om),
    function(nm) rnorm(n, 0, sqrt(om[nm, nm])),
    numeric(n)
  ))
}

etas_nihss <- draw_etas(ui_nihss, N_SUBJ)
etas_bi    <- draw_etas(ui_bi,    N_SUBJ)
```

# Trajectory simulation

A `$PRED` transition model does not produce a trajectory by itself: the score at
each visit is an input to the next visit's prediction. The authors generated
their own 10,000-subject diagnostic simulation by walking the hierarchy visit by
visit, and that is what the loop below does -- calling the packaged model once
per visit, drawing the Bernoulli events and the logit-scale residual, and
carrying the reconstructed score forward.

```{r simulate-scale}
# One scale's trajectory. `cfg` names the scale-specific columns and constants
# so the same walk serves both models.
simulate_scale <- function(ui, cohort, etas, visits, cfg) {
  state <- tibble::tibble(
    id      = cohort$id,
    score   = cfg$start_score,
    dropped = FALSE
  )

  out <- tibble::tibble(
    id = state$id, day = cfg$start_day, score = state$score,
    dropped = FALSE, event = "start"
  )

  for (k in seq_len(nrow(visits))) {
    ev <- cohort
    ev$time      <- visits$day[k]
    ev$evid      <- 0
    ev$amt       <- NA_real_
    ev$T_PREVOBS <- visits$t_prevobs[k]
    ev[[cfg$prev_col]] <- state$score
    ev <- dplyr::bind_cols(ev, etas)

    sim <- as.data.frame(rxode2::rxSolve(ui, ev, omega = NA))
    stopifnot(nrow(sim) == nrow(state))

    n <- nrow(sim)
    # Residual error enters INSIDE the logit, before the expit -- the source
    # writes IM = ... + THETA(6)*EPS(1) then IMZ = exp(IM)/(1+exp(IM)).
    imz <- plogis(sim[[cfg$logit_improve]] + rnorm(n, 0, cfg$sd_imz))
    dez <- plogis(sim[[cfg$logit_decline]] + rnorm(n, 0, cfg$sd_dez))

    improved <- runif(n) < sim$p_improve
    not_max  <- runif(n) < sim$p_notmax
    drops    <- runif(n) < sim$p_dropout

    new_score <- cfg$reconstruct(state$score, imz, dez, improved, not_max)
    event <- ifelse(
      improved,
      ifelse(not_max, "improved", "recovered"),
      ifelse(drops, "dropout", "declined")
    )

    # A subject who dropped out at an earlier visit contributes no further
    # score and no further event; the state is absorbing.
    was_dropped   <- state$dropped
    state$score   <- ifelse(was_dropped, state$score, new_score)
    state$dropped <- was_dropped | (!improved & drops)

    out <- dplyr::bind_rows(out, tibble::tibble(
      id      = state$id,
      day     = visits$day[k],
      score   = ifelse(state$dropped, NA_real_, state$score),
      dropped = state$dropped,
      event   = ifelse(was_dropped, "dropped earlier", event)
    ))
  }
  out
}
```

```{r reconstruct-rules}
# NIHSS: 0 is full recovery; an improvement moves the score toward 0.
reconstruct_nihss <- function(prev, imz, dez, improved, not_max) {
  ifelse(
    improved,
    ifelse(not_max, (1 - imz) * (prev + 1), 0),
    dez * (42 - prev) + prev
  )
}

# Barthel Index: 100 is full recovery; an improvement moves the score toward 100.
reconstruct_bi <- function(prev, imz, dez, improved, not_max) {
  ifelse(
    improved,
    ifelse(not_max, prev + imz * (100 - prev), 100),
    (1 - dez) * (prev + 1)
  )
}
```

```{r run-simulation}
th <- function(ui, nm) ui$theta[[nm]]

# NIHSS: admission, then days 7, 30 and 90.
traj_nihss <- simulate_scale(
  ui_nihss, cohort, etas_nihss,
  visits = tibble::tibble(day = c(7, 30, 90), t_prevobs = c(7, 23, 60)),
  cfg = list(
    start_day      = 0,
    start_score    = cohort$SCORE_NIHSS,
    prev_col       = "SCORE_NIHSS_PREV",
    logit_improve  = "logitimpr",
    logit_decline  = "logitdecl_i",
    sd_imz         = th(ui_nihss, "addSd_imz"),
    sd_dez         = th(ui_nihss, "addSd_dez"),
    reconstruct    = reconstruct_nihss
  )
)

# Barthel Index: no admission assessment, so the walk starts at the day-7 score.
# The paper reports no day-7 BI distribution, so it is seeded from each
# subject's baseline NIHSS on the reversed scale (see Assumptions and deviations).
bi_day7 <- round(pmax(0, pmin(100, (42 - cohort$SCORE_NIHSS) / 42 * 100)) / 5) * 5

traj_bi <- simulate_scale(
  ui_bi, cohort, etas_bi,
  visits = tibble::tibble(day = c(30, 60, 90), t_prevobs = c(23, 30, 30)),
  cfg = list(
    start_day      = 7,
    start_score    = bi_day7,
    prev_col       = "SCORE_BI_PREV",
    logit_improve  = "logitimpr_i",
    logit_decline  = "logitdecl_i",
    sd_imz         = th(ui_bi, "addSd_imz"),
    sd_dez         = th(ui_bi, "addSd_dez"),
    reconstruct    = reconstruct_bi
  )
)
```

# Replicate published figures

## Figure 3 -- simulated score distributions over time

Karlsson 2010 Figure 3 plots simulated scores against the observed data for
each scale, and the Results state that "50th and 90th percentiles of simulated
scores match the corresponding observed score percentiles for both scales". The
observed overlay is a figure panel with no tabulated values in the article or
either supplement, so only the simulated side is reproducible here.

```{r figure-3, fig.width = 7, fig.height = 3.5}
pctl <- function(traj, scale_label) {
  traj |>
    dplyr::filter(!is.na(score)) |>
    dplyr::group_by(day) |>
    dplyr::summarise(
      Q10 = quantile(score, 0.10),
      Q50 = quantile(score, 0.50),
      Q90 = quantile(score, 0.90),
      .groups = "drop"
    ) |>
    dplyr::mutate(scale = scale_label)
}

fig3 <- dplyr::bind_rows(
  pctl(traj_nihss, "a  NIHSS (0-42, lower is better)"),
  pctl(traj_bi,    "b  Barthel Index (0-100, higher is better)")
)

ggplot(fig3, aes(day, Q50)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  facet_wrap(~scale, scales = "free_y") +
  labs(x = "Day since stroke", y = "Score",
       title = "Figure 3 - simulated score percentiles (10th, 50th, 90th)",
       caption = "Replicates the simulated side of Figure 3 of Karlsson 2010.")
```

## Figure 4 -- simulated change from baseline

```{r figure-4, fig.width = 7, fig.height = 3.5}
cfb <- function(traj, scale_label) {
  base <- traj |>
    dplyr::filter(day == min(day)) |>
    dplyr::select(id, base_score = score)
  traj |>
    dplyr::inner_join(base, by = "id") |>
    dplyr::filter(!is.na(score), day > min(day)) |>
    dplyr::mutate(delta = score - base_score) |>
    dplyr::group_by(day) |>
    dplyr::summarise(
      Q10 = quantile(delta, 0.10),
      Q50 = quantile(delta, 0.50),
      Q90 = quantile(delta, 0.90),
      .groups = "drop"
    ) |>
    dplyr::mutate(scale = scale_label)
}

fig4 <- dplyr::bind_rows(
  cfb(traj_nihss, "a  NIHSS change from admission"),
  cfb(traj_bi,    "b  Barthel Index change from day 7")
)

ggplot(fig4, aes(day, Q50)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  facet_wrap(~scale, scales = "free_y") +
  labs(x = "Day since stroke", y = "Change from baseline score",
       title = "Figure 4 - simulated change from baseline (10th, 50th, 90th percentiles)",
       caption = "Replicates the simulated side of Figure 4 of Karlsson 2010.")
```

## Transition-event mix and informative dropout

The feature that distinguishes this model class from a conventional
longitudinal analysis is that dropout is modelled rather than discarded, and is
driven by the predicted score. The panel below shows the event mix per visit and
the dependence of the dropout probability on the predicted declined score.

```{r event-mix, fig.width = 7, fig.height = 3.5}
event_mix <- dplyr::bind_rows(
  traj_nihss |> dplyr::filter(event != "start") |> dplyr::mutate(scale = "NIHSS"),
  traj_bi    |> dplyr::filter(event != "start") |> dplyr::mutate(scale = "Barthel Index")
) |>
  dplyr::count(scale, day, event) |>
  dplyr::group_by(scale, day) |>
  dplyr::mutate(fraction = n / sum(n)) |>
  dplyr::ungroup()

ggplot(event_mix, aes(factor(day), fraction, fill = event)) +
  geom_col() +
  facet_wrap(~scale) +
  labs(x = "Day since stroke", y = "Fraction of subjects", fill = NULL,
       title = "Transition-event mix at each visit")
```

```{r informative-dropout, fig.width = 6, fig.height = 3.2}
# Deterministic covariate sweeps at typical values -- one row per grid point,
# not per simulated participant.
grid_n <- solve_nihss(seq(0, 42, length.out = 100))
grid_b <- solve_bi(seq(0, 100, length.out = 100))

dplyr::bind_rows(
  tibble::tibble(scale = "NIHSS (0-42)",
                 predicted = grid_n$nihss_decline, p_dropout = grid_n$p_dropout),
  tibble::tibble(scale = "Barthel Index (0-100)",
                 predicted = grid_b$bi_decline,    p_dropout = grid_b$p_dropout)
) |>
  ggplot(aes(predicted, p_dropout)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~scale, scales = "free_x") +
  labs(x = "Model-predicted declined score", y = "p(dropout)",
       title = "Informative dropout: p(dropout) against the predicted declined score",
       subtitle = "Typical values, day 30. Opposite slopes because the scales run in opposite directions.")
```

# Validation

This paper has no PK layer, no dose and no concentration, so NCA is not the
applicable check. The gates below are the structural and behavioural ones the
model class admits.

```{r validation}
score_range <- function(traj) range(traj$score, na.rm = TRUE)
r_nihss <- score_range(traj_nihss)
r_bi    <- score_range(traj_bi)

# Median score at the day-90 endpoint versus baseline.
med_at <- function(traj, d) median(traj$score[traj$day == d], na.rm = TRUE)
nihss_base <- med_at(traj_nihss, 0)
nihss_d90  <- med_at(traj_nihss, 90)
bi_base    <- med_at(traj_bi, 7)
bi_d90     <- med_at(traj_bi, 90)

# Fraction still in study at day 90.
retained <- function(traj) mean(!traj$dropped[traj$day == max(traj$day)])
ret_nihss <- retained(traj_nihss)
ret_bi    <- retained(traj_bi)

checks <- tibble::tribble(
  ~Check, ~Value, ~Requirement,
  "NIHSS scores stay within [0, 43]",
    sprintf("[%.2f, %.2f]", r_nihss[1], r_nihss[2]), "the 0-42 scale plus the source's +1 offset",
  "Barthel Index scores stay within [0, 101]",
    sprintf("[%.2f, %.2f]", r_bi[1], r_bi[2]),       "the 0-100 scale plus the source's +1 offset",
  "Median NIHSS improves from admission to day 90",
    sprintf("%.1f -> %.1f", nihss_base, nihss_d90),  "lower is better",
  "Median Barthel Index improves from day 7 to day 90",
    sprintf("%.1f -> %.1f", bi_base, bi_d90),        "higher is better",
  "Fraction retained at day 90 (NIHSS model)",
    sprintf("%.2f", ret_nihss),                      "between 0.50 and 0.95",
  "Fraction retained at day 90 (BI model)",
    sprintf("%.2f", ret_bi),                         "between 0.60 and 0.99"
)

knitr::kable(checks, caption = "Structural and behavioural validation of the packaged models.")
```

```{r validation-gate}
stopifnot(
  # Bounds are a hard property of the reconstruction rules, not a cohort draw.
  # The +1 upper slack is the source's own offset: an "improvement" from
  # SCORE_NIHSS_PREV can land at SCORE_NIHSS_PREV + 1 when imz is near 0, and a
  # "decline" from a maximal Barthel Index can land at 101 when dez is near 0.
  r_nihss[1] >= 0, r_nihss[2] <= 43,
  r_bi[1]    >= 0, r_bi[2]    <= 101,

  # Direction and magnitude of the population trend at the endpoint. Asserted
  # as a trend with headroom rather than as a step-by-step ordering, because
  # visit-to-visit medians of a cohort this size can cross. Realised on the
  # authoring run: NIHSS median 17.0 -> 9.3 (a 7.7-point improvement) and BI
  # median 60.0 -> 99.6 (39.6 points). The bounds below sit well inside that
  # and still go red if the improvement and decline reconstruction rules are
  # swapped, which would send both trends the other way.
  nihss_base - nihss_d90 > 2,
  bi_d90 - bi_base       > 10,

  # Retention at day 90. This is the gate on the informative-dropout sign: with
  # slope_score_p_dropout flipped, the NIHSS model drops essentially nobody
  # (retention -> 1.0) and the BI model drops most of the cohort. Realised on
  # the authoring run: 0.76 (NIHSS) and 0.92 (BI); binomial standard errors at
  # n = 200 are 0.03 and 0.02, so both bounds carry several standard errors of
  # headroom in each direction.
  ret_nihss > 0.50, ret_nihss < 0.95,
  ret_bi    > 0.60, ret_bi    < 0.99
)
```

A note on the bounds check: the `+ 1` slack is not a tolerance, it is the
source's convention. Karlsson 2010's Table I treats an unchanged score as a
decline, and the reconstruction rules carry a `+ 1` offset
(`(1 - imz) * (SCORE_NIHSS_PREV + 1)`, `(1 - dez) * (SCORE_BI_PREV + 1)`) so
that a full improvement can reach the scale endpoint exactly. The cost is that a
minimal improvement can overshoot by up to one point. The packaged models
reproduce that rather than clamping it.

# Assumptions and deviations

**Parameter provenance**

- **Supplement A over Supplement B.** Every packaged value is the final estimate
  from Supplement A. Supplement B's `$THETA` / `$OMEGA` records are NONMEM
  initial estimates and disagree with Supplement A on several NIHSS parameters
  and on 11 of 15 BI thetas plus the omega (1.25 vs 2.08). The adjudication is
  the paper's own stated NIHSS age effect, reproduced under Supplement A and not
  under Supplement B; see the *Source trace* section.
- **BI age effect (claim 3) is a recorded deviation**, reproducing as
  `r sprintf("%.4f", claim3)` against a published 0.07. Not tuned.

**Corrections to the source supplements**

- **`TLS` is mislabelled in the Supplement B dataset legend.** The legend reads
  "TLS = Time since start of study (i.e. time since stroke) in days". The
  tabulated data falsify it: in the NIHSS excerpt subject 2 has
  `DAY 8 -> TLS 7`, `DAY 33 -> TLS 25` (= 33 - 8), `DAY 62 -> TLS 29` (= 62 - 33);
  in the BI excerpt `TLS` is non-monotonic in `DAY`, which a time-since-stroke
  clock cannot be. `TLS` is time since the **previous observation** and is
  packaged as `T_PREVOBS`. Table II corroborates this by listing "Influence of
  time since previous observation" and "Influence of time since baseline" as
  separate rows; the second clock is `TSS = DAY`, carried here by rxode2's
  `time`. Taking the legend at face value would silently mis-specify
  `e_t_prevobs_p_notmax` (NIHSS) and `e_t_prevobs_decl` (BI).
- **Four undefined variables in the BI control stream are omitted.** That stream
  writes `DECOV = DETLS + DEGGT + DEANG`, `A1COV = A1PBI + A1AGE + A1CGL` and
  `A2COV = A2PBI + A2TLS + A2TSS`, but `DEGGT`, `DEANG`, `A1CGL` and `A2TLS` are
  never assigned anywhere in it -- vestigial covariate-screening terms, with GGT,
  ANG and CGL surviving undropped in the BI `$INPUT`. None has a counterpart
  among Supplement A's 15 BI thetas or in Table II, so all four are zero in the
  final model and are omitted from the packaged file.
- **The BI `$THETA(1)` inline comment is wrong.** It reads "Probability of
  decline (logit scale)", but the code assigns `P1A = P1 ; Improvement`,
  Supplement A names the parameter `PrIM`, and the covariate signs match
  improvement. The code and the paper agree; the comment does not.
- **No IIV on the BI improvement magnitude**, despite the Results prose. Karlsson
  2010 states the BI relative-improvement model included "terms for
  interindividual variability and residual variability", but the Supplement B
  stream has `IM = TVIM + RV` with no `ETA` term while `DE = TVDE + ETA(1) + RV`,
  and Supplement A lists a single omega labelled "IIV on decline". The equation
  governs.
- **`AGE = -99` missing-value branching is not reproduced.** The BI stream
  branches on `IF(AGE.EQ.-99)` to zero both age effects. That is a NONMEM
  data-coding convention, not model structure; supply a real age.

**Encoding**

- **Placeholder residuals on the three probability endpoints.** The source
  likelihood for `p_improve`, `p_notmax` and `p_dropout` is Bernoulli on the
  observed binary event (NONMEM `F_FLAG = 1`), which carries no residual-error
  parameter. rxode2 requires an error model per endpoint, so each probability is
  emitted with a fixed additive residual SD of 0.001, following
  `Fukae_2024_valemetostat_orr_investigator`. Do not read these as estimates.
- **`addSd_imz` and `addSd_dez` are one source parameter written twice.** The
  source declares `$SIGMA 1 FIX` with `RV = THETA(6)*EPS(1)` and applies that
  single term to both the improvement and the decline function. rxode2 requires
  a distinct endpoint parameter per endpoint.
- **The etas are not mu-referenced.** Which eta applies switches with time in
  the NIHSS model, so the eta cannot sit on a simple `theta + eta` line. rxode2
  warns about this at parse time; it is correct for this model and affects
  simulation not at all.

**Simulation choices in this vignette**

- **Covariate dispersions are assumed.** The paper reports means and ranges but
  no standard deviations. Age is drawn N(71.7, 12) truncated to 26-90 and
  baseline NIHSS N(16.8, 5) truncated to 4-31, so that the simulated means and
  ranges match the published ones.
- **The day-7 Barthel Index seed is constructed, not published.** The BI has no
  admission assessment and the paper reports no day-7 BI distribution, so each
  subject's starting BI is mapped from their baseline NIHSS on the reversed
  scale and rounded to the scale's five-point grid. This is a starting condition
  for the demonstration only; it is not a model parameter and nothing in the
  packaged model depends on it.
- **`T_PREVOBS` at the first NIHSS transition is taken as 7 days** (admission at
  day 0). The source dataset implies an admission assessment on day 1
  (`DAY 7 -> TLS 6`), a one-day difference with no material effect.
- **The recursion draws Bernoulli events and logit-scale residuals in R.** The
  packaged models supply every probability, every logit-scale prediction and
  both reconstruction rules; only the random draws and the carry-forward of the
  previous score happen in the vignette. This is the same walk the authors used
  to generate their 10,000-subject diagnostic simulation, at the 200-subject
  scale this package caps cohorts at.
