---
title: "Amikacin (Kato 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Amikacin (Kato 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r meta}
mod <- readModelDb("Kato_2021_amikacin")
mod_meta <- rxode2::rxode2(mod)$meta
```

* Citation: `r mod_meta$reference`
* Description: `r mod_meta$description`
* Article (DOI): <https://doi.org/10.3390/antibiotics10020100>

Kato 2021 describes amikacin in 15 hospitalized patients aged 70 years and
over with a one-compartment model in which clearance is directly proportional
to Cockcroft-Gault creatinine clearance (CCr). The model was estimated with the
non-parametric adaptive grid (NPAG) algorithm of Pmetrics. The paper then uses
Monte Carlo simulation to screen 200-2000 mg regimens given every 24, 48 or 72
h against an efficacy target (Cmax/MIC >= 8) and a safety target (Cmin < 4
mg/L). This vignette checks the packaged model against the paper's Table 2
parameter estimates, the per-kilogram values quoted in its Discussion, and the
1080-cell probability-of-target-attainment grid of Table 3.

# Population

The analysis is a single-centre retrospective therapeutic drug monitoring study
at Aichi Medical University Hospital, Japan (September 2009 - February 2015).
Fifteen patients aged 71-95 years (mean 80.6) were included, 9 female and 6
male. They were light (mean weight 44.8 kg, range 32.5-67.3; mean BMI 19.1) and
hypoalbuminaemic (mean albumin 2.5 g/dL). Mean Cockcroft-Gault creatinine
clearance was 52.9 mL/min (median 52.1, range 10.9-94.9). Infections were
mostly pneumonia and bacteremia, and *Pseudomonas aeruginosa* was the commonest
isolate. Amikacin was given at 200-1000 mg/day (mean 440 mg/day, 9.6 mg/kg/day)
by 0.5-1.0 h infusion, and 33 serum concentrations (2-3 per patient, a peak and
a trough) were measured (Kato 2021 Table 1, Sections 2.1-2.2 and 4.1-4.2).
Patients on renal replacement therapy were excluded.

```{r population}
str(mod_meta$population)
```

# Source trace

| Element | Value | Source location |
|---|---|---|
| Structure: one compartment, first-order elimination, IV infusion | -- | Section 2.2 ('the one-compartment linear model described the data adequately'); Section 4.3 |
| `lcl` (CLs at CCr = 52.9 mL/min) | log(2.25) L/h | Table 2, CL mean 2.25 (SD 0.78, median 2.19, CV 34.6%); abstract |
| `lvc` (Vd) | log(18.0) L | Table 2, V mean 18.0 (SD 3.4, median 17.1, CV 18.9%); abstract |
| CL covariate model `CL = CLs * (CRCL / 52.9)` | -- | Section 2.2, final-model equation |
| `etalcl` | log(0.346^2 + 1) = 0.113075 | Table 2, CL CV 34.6% (log-normal approximation of the NPAG distribution) |
| `etalvc` | log(0.189^2 + 1) = 0.035098 | Table 2, V CV 18.9% (log-normal approximation) |
| `propSd`, `addSd` | fixed(0) | Not reported; see Assumptions and deviations |
| Covariate `CRCL` (raw Cockcroft-Gault, mL/min) | -- | Section 4.2; Table 1 footnote |

# Parameter checks against the paper

The paper's Table 2 reports each parameter's mean, SD and CV%. `CV% = SD /
mean` holds for both rows, which confirms the CV% column describes the spread
of the population distribution (a between-subject statistic) rather than
estimation precision. The Discussion quotes the estimates per kilogram at the
cohort mean weight of 44.8 kg as 0.05 L/h/kg and 0.40 L/kg.

```{r param-checks}
ini_df <- rxode2::rxode2(mod)$iniDf
theta <- setNames(ini_df$est, ini_df$name)
cl_typ <- exp(theta[["lcl"]])
v_typ <- exp(theta[["lvc"]])

param_check <- data.frame(
  quantity = c("CV% of CL = SD / mean", "CV% of V = SD / mean",
               "CL per kg (L/h/kg)", "V per kg (L/kg)"),
  reproduced = c(100 * 0.78 / 2.25, 100 * 3.4 / 18.0,
                 cl_typ / 44.8, v_typ / 44.8),
  published = c(34.6, 18.9, 0.05, 0.40)
)
knitr::kable(param_check, digits = 3,
             caption = "Table 2 internal consistency and Discussion per-kg values.")

stopifnot(
  abs(param_check$reproduced[1] - 34.6) < 0.2,
  abs(param_check$reproduced[2] - 18.9) < 0.2,
  abs(round(cl_typ / 44.8, 2) - 0.05) < 1e-9,
  abs(round(v_typ / 44.8, 2) - 0.40) < 1e-9
)
```

# Virtual cohort

The paper does not publish individual data, so a 200-subject cohort is drawn
with creatinine clearance from a normal distribution matching the Table 1 mean
and SD (52.9 +/- 22.8 mL/min), rejecting and redrawing values outside the
observed range 10.9-94.9 mL/min. Between-subject random effects are drawn in
base R from the model's own `omega`, so the cohort is identical on every rxode2
build, and they are passed to the typical-value model as data columns.

```{r cohort}
omega <- rxode2::rxode2(mod)$omega
mod_typ <- rxode2::zeroRe(mod)

draw_crcl <- function(n, mean = 52.9, sd = 22.8, lo = 10.9, hi = 94.9) {
  out <- numeric(0)
  while (length(out) < n) {
    x <- rnorm(2 * n, mean, sd)
    out <- c(out, x[x >= lo & x <= hi])
  }
  out[seq_len(n)]
}

set.seed(2021)
n_sub <- 200
cohort <- data.frame(
  id = seq_len(n_sub),
  CRCL = draw_crcl(n_sub),
  etalcl = rnorm(n_sub, 0, sqrt(omega["etalcl", "etalcl"])),
  etalvc = rnorm(n_sub, 0, sqrt(omega["etalvc", "etalvc"]))
)
summary(cohort$CRCL)
```

# Simulation

Each virtual patient receives a single 400 mg dose (the Table 1 median daily
dose) as a 0.5 h infusion (the Table 1 median infusion time), observed on a
dense grid out to 48 h, from which PKNCA extrapolates to infinity. The
grid stops at 48 h, and the solver tolerances are tightened, because
fast-clearing subjects otherwise fall to numerical-noise concentrations
(including tiny negative values) that PKNCA cannot use.

```{r simulate}
obs_times <- sort(unique(c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 12,
                           16, 24, 30, 36, 42, 48)))
dose_rows <- cohort |>
  mutate(time = 0, amt = 400, rate = 400 / 0.5, evid = 1L, cmt = "central")
obs_rows <- tidyr::crossing(cohort, time = obs_times) |>
  mutate(amt = 0, rate = 0, evid = 0L, cmt = "central")
events <- bind_rows(dose_rows, obs_rows) |>
  arrange(id, time, desc(evid)) |>
  select(id, time, amt, rate, evid, cmt, CRCL, etalcl, etalvc)

sim <- rxode2::rxSolve(mod_typ, events = events, returnType = "data.frame",
                       keep = "CRCL", atol = 1e-12, rtol = 1e-10)
stopifnot(all(sim$Cc >= 0))
sim_obs <- sim |> filter(!is.na(Cc))
```

## Solver check against the closed-form solution

With the random effects supplied as data, each simulated profile must equal
the analytical one-compartment infusion solution for that subject's own
clearance and volume. Both sides use the same drawn parameters, so the only
difference is numerical error and the bound is tight.

```{r closed-form}
closed_form <- function(t, dose, tinf, cl, v) {
  k <- cl / v
  r0 <- dose / tinf
  ifelse(t <= tinf,
         r0 / cl * (1 - exp(-k * t)),
         r0 / cl * (1 - exp(-k * tinf)) * exp(-k * (t - tinf)))
}
chk <- sim_obs |>
  select(id, time, Cc, CRCL) |>
  left_join(cohort |> select(id, etalcl, etalvc), by = "id") |>
  mutate(
    cl_i = cl_typ * exp(etalcl) * CRCL / 52.9,
    v_i = v_typ * exp(etalvc),
    Cc_cf = closed_form(time, 400, 0.5, cl_i, v_i)
  ) |>
  # Relative error is compared above 0.01 mg/L (far below the 0.8 mg/L assay
  # limit), where the solver's absolute tolerance does not dominate.
  filter(time > 0, Cc_cf > 0.01)
max_rel_err <- max(abs(chk$Cc - chk$Cc_cf) / chk$Cc_cf)
max_rel_err
stopifnot(max_rel_err < 1e-4)
```

## Simulated profiles

```{r profiles}
sim_obs |>
  mutate(crcl_band = cut(CRCL, c(0, 30, 60, 100),
                         labels = c("CRCL < 30", "CRCL 30-60", "CRCL > 60"))) |>
  group_by(crcl_band, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  filter(time > 0) |>
  ggplot(aes(time, Q50, colour = crcl_band, fill = crcl_band)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.15, colour = NA) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Time after start of infusion (h)",
       y = "Amikacin serum concentration (mg/L)",
       colour = NULL, fill = NULL,
       title = "Single 400 mg dose over 0.5 h, 200 virtual patients",
       subtitle = "Median and 5th-95th percentiles by creatinine-clearance band") +
  theme_minimal()
```

# PKNCA validation

The paper reports no non-compartmental summary, so the NCA is checked
against the model's own identity: for linear IV kinetics, `AUC0-inf = Dose /
CL` for each subject's individual clearance.

```{r pknca}
nca_conc <- sim_obs |>
  mutate(treatment = "400 mg single dose") |>
  select(id, time, Cc, treatment) |>
  as.data.frame()
nca_dose <- dose_rows |>
  mutate(treatment = "400 mg single dose") |>
  select(id, time, amt, treatment) |>
  as.data.frame()

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ time | treatment + id,
                             concu = "mg/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(nca_dose, amt ~ time | treatment + id,
                             doseu = "mg")
intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
knitr::kable(summary(nca_res), caption = "Simulated single-dose NCA (PKNCA).")

auc_chk <- as.data.frame(nca_res) |>
  filter(PPTESTCD == "aucinf.obs") |>
  mutate(id = as.integer(id)) |>
  left_join(cohort, by = "id") |>
  mutate(
    cl_i = cl_typ * exp(etalcl) * CRCL / 52.9,
    pct_diff = 100 * (PPORRES - 400 / cl_i) / (400 / cl_i)
  )
summary(auc_chk$pct_diff)
stopifnot(
  abs(median(auc_chk$pct_diff)) < 2,
  quantile(abs(auc_chk$pct_diff), 0.9) < 5
)
```

# Replicate Table 3: probability of target attainment

Table 3 of Kato 2021 marks, for each regimen (200-2000 mg every 24, 48 or 72
h) and creatinine clearance (10-90 mL/min), whether at least 90% of 1000
simulated patients met each target: `Cmin < 4 mg/L` and `Cmax/MIC >= 8` for
MICs of 4, 8 and 16 mg/L. That is 3 x 9 x 10 x 4 = 1080 binary cells. The grid
below was transcribed by the maintainers from the three raster panels of
Table 3, by classifying each printed glyph as `+` or `-`.

Section 4.6 of the paper defines Cmax as the concentration 1 h after the start
of a 30 min infusion and Cmin as the concentration 23.5, 47.5 or 71.5 h after
the start of infusion for the q24h, q48h and q72h regimens. It does not say
which dose the Cmax refers to. The primary reading below takes both from the
first dose; a sensitivity reading takes Cmax after the tenth dose, which is
close to steady state.

```{r table3-paper}
t3_rows <- list(
  q24h = c(
    "---- ---- -+-- -+-- -++- -++- -++- -++- -++- -+++",
    "---- ---- -+-- -+-- -+-- -++- -++- -++- -++- -++-",
    "+--- ---- -+-- -+-- -+-- -++- -++- -++- -++- -++-",
    "+--- ---- ---- -+-- -+-- -+-- -++- -++- -++- -++-",
    "+--- ---- ---- -+-- -+-- -+-- -++- -++- -++- -++-",
    "+--- +--- ---- -+-- -+-- -+-- -++- -++- -++- -++-",
    "+--- +--- +--- -+-- -+-- -+-- -+-- -++- -++- -++-",
    "+--- +--- +--- ++-- -+-- -+-- -+-- -++- -++- -++-",
    "+--- +--- +--- ++-- ++-- -+-- -+-- -++- -++- -++-"
  ),
  q48h = c(
    "---- ---- -+-- -+-- -+-- -++- -++- -++- -++- -++-",
    "+--- ---- ---- -+-- -+-- -+-- -++- -++- -++- -++-",
    "+--- +--- ---- -+-- -+-- -+-- -++- -++- -++- -++-",
    "+--- +--- +--- ++-- -+-- -+-- -+-- -++- -++- -++-",
    "+--- +--- +--- ++-- ++-- ++-- ++-- -++- -++- -++-",
    "+--- +--- +--- ++-- ++-- ++-- ++-- ++-- +++- +++-",
    "+--- +--- +--- ++-- ++-- ++-- ++-- ++-- +++- +++-",
    "+--- +--- +--- +--- ++-- ++-- ++-- ++-- +++- +++-",
    "+--- +--- +--- +--- ++-- ++-- ++-- ++-- +++- +++-"
  ),
  q72h = c(
    "+--- ---- -+-- -+-- -+-- -++- -++- -++- -++- -++-",
    "+--- +--- ---- -+-- -+-- -+-- -++- -++- -++- -++-",
    "+--- +--- +--- ++-- ++-- -+-- -+-- -++- -++- -++-",
    "+--- +--- +--- ++-- ++-- ++-- ++-- ++-- +++- +++-",
    "+--- +--- +--- +--- ++-- ++-- ++-- ++-- +++- +++-",
    "+--- +--- +--- +--- ++-- ++-- ++-- ++-- +++- +++-",
    "+--- +--- +--- +--- ++-- ++-- ++-- ++-- +++- +++-",
    "+--- +--- +--- +--- ++-- ++-- ++-- ++-- +++- +++-",
    "+--- +--- +--- +--- ++-- ++-- ++-- ++-- +++- +++-"
  )
)
targets <- c("cmin", "mic4", "mic8", "mic16")
paper_t3 <- bind_rows(lapply(names(t3_rows), function(iv) {
  bind_rows(lapply(seq_along(t3_rows[[iv]]), function(i) {
    glyphs <- strsplit(gsub(" ", "", t3_rows[[iv]][i]), "")[[1]]
    stopifnot(length(glyphs) == 40)
    data.frame(
      interval = iv,
      crcl = 10 * i,
      dose = rep(seq(200, 2000, 200), each = 4),
      target = rep(targets, times = 10),
      paper = as.integer(glyphs == "+")
    )
  }))
}))
stopifnot(nrow(paper_t3) == 1080)
```

The 1080 cells are simulated with 200 patients per interval and creatinine
clearance (the paper used 1000), sharing one set of base-R random-effect
draws. Because the model is linear, every regimen is a dose multiple of a
single 1000 mg solve.

```{r table3-sim}
n_pta <- 200
set.seed(20210120)
pta_eta <- data.frame(
  sub = seq_len(n_pta),
  etalcl = rnorm(n_pta, 0, sqrt(omega["etalcl", "etalcl"])),
  etalvc = rnorm(n_pta, 0, sqrt(omega["etalvc", "etalvc"]))
)
pta_design <- tidyr::crossing(tau = c(24, 48, 72), crcl = seq(10, 90, 10),
                              pta_eta) |>
  mutate(id = row_number())

pta_dose <- pta_design |>
  mutate(time = 0, amt = 1000, rate = 1000 / 0.5, ii = tau, addl = 9L,
         evid = 1L)
pta_obs <- bind_rows(
  pta_design |> mutate(time = 1, what = "cmax_first"),
  pta_design |> mutate(time = tau - 0.5, what = "cmin"),
  pta_design |> mutate(time = 9 * tau + 1, what = "cmax_10th")
) |>
  mutate(amt = 0, rate = 0, ii = 0, addl = 0L, evid = 0L)
pta_events <- bind_rows(pta_dose, pta_obs) |>
  mutate(cmt = "central", CRCL = crcl) |>
  arrange(id, time, desc(evid)) |>
  select(id, time, amt, rate, ii, addl, evid, cmt, CRCL, etalcl, etalvc)

pta_sim <- rxode2::rxSolve(mod_typ, events = pta_events,
                           returnType = "data.frame") |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc) |>
  left_join(pta_obs |> select(id, time, what, tau, crcl), by = c("id", "time"))
stopifnot(nrow(pta_sim) == 3 * nrow(pta_design))

pta_wide <- pta_sim |>
  select(id, tau, crcl, what, Cc) |>
  tidyr::pivot_wider(names_from = what, values_from = Cc)

pta_grid <- function(wide, cmax_col) {
  bind_rows(lapply(seq(200, 2000, 200), function(d) {
    wide |>
      mutate(cmax = .data[[cmax_col]] * d / 1000,
             cmin_d = cmin * d / 1000) |>
      group_by(tau, crcl) |>
      summarise(cmin = mean(cmin_d < 4), mic4 = mean(cmax / 4 >= 8),
                mic8 = mean(cmax / 8 >= 8), mic16 = mean(cmax / 16 >= 8),
                .groups = "drop") |>
      mutate(dose = d)
  })) |>
    tidyr::pivot_longer(c(cmin, mic4, mic8, mic16), names_to = "target",
                        values_to = "pta") |>
    mutate(interval = paste0("q", tau, "h"), sim = as.integer(pta >= 0.9)) |>
    select(interval, crcl, dose, target, pta, sim)
}
cmp_first <- inner_join(paper_t3, pta_grid(pta_wide, "cmax_first"),
                        by = c("interval", "crcl", "dose", "target"))
cmp_10th <- inner_join(paper_t3, pta_grid(pta_wide, "cmax_10th"),
                       by = c("interval", "crcl", "dose", "target"))
stopifnot(nrow(cmp_first) == 1080, nrow(cmp_10th) == 1080)
```

```{r table3-agreement}
agree_tab <- function(cmp, label) {
  cmp |>
    group_by(target) |>
    summarise(agreement = mean(sim == paper), .groups = "drop") |>
    bind_rows(data.frame(target = "all", agreement = mean(cmp$sim == cmp$paper))) |>
    mutate(reading = label)
}
agreement <- bind_rows(
  agree_tab(cmp_first, "Cmax after the first dose"),
  agree_tab(cmp_10th, "Cmax after the tenth dose")
)
agreement |>
  tidyr::pivot_wider(names_from = reading, values_from = agreement) |>
  dplyr::rename("Target" = target) |>
  knitr::kable(digits = 3,
               caption = "Fraction of Table 3 cells whose >= 90% PTA flag is reproduced.")
```

```{r table3-figure, fig.height = 7}
cmp_first |>
  mutate(
    status = case_when(
      sim == paper & paper == 1 ~ "both >= 90%",
      sim == paper ~ "both < 90%",
      sim == 1 ~ "simulated only",
      TRUE ~ "paper only"
    ),
    target = factor(target, targets,
                    c("Cmin < 4", "Cmax/MIC >= 8, MIC 4",
                      "Cmax/MIC >= 8, MIC 8", "Cmax/MIC >= 8, MIC 16"))
  ) |>
  ggplot(aes(factor(dose), factor(crcl), fill = status)) +
  geom_tile(colour = "white") +
  facet_grid(target ~ interval) +
  scale_fill_manual(values = c("both >= 90%" = "#2c7fb8",
                               "both < 90%" = "#d9d9d9",
                               "simulated only" = "#fdae61",
                               "paper only" = "#d7191c")) +
  labs(x = "Dose (mg)", y = "Creatinine clearance (mL/min)", fill = NULL,
       title = "Replicates Table 3 of Kato 2021 (first-dose reading)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 6),
        axis.text.y = element_text(size = 6),
        legend.position = "bottom")
```

The two readings agree with the paper equally well on the trough target,
since the trough is always taken after the first dose. For the Cmax targets
the readings differ mainly at q24h, the only interval with appreciable
accumulation, where the paper's flags sit closer to the tenth-dose reading:

```{r cmax-by-interval}
cmax_by_interval <- bind_rows(
  cmp_first |> mutate(reading = "first dose"),
  cmp_10th |> mutate(reading = "tenth dose")
) |>
  filter(target != "cmin") |>
  group_by(interval, reading) |>
  summarise(agreement = mean(sim == paper), .groups = "drop") |>
  tidyr::pivot_wider(names_from = reading, values_from = agreement)
cmax_by_interval |>
  dplyr::rename("Interval" = interval,
                "Cmax agreement, first dose" = `first dose`,
                "Cmax agreement, tenth dose" = `tenth dose`) |>
  knitr::kable(digits = 3, caption = "Agreement on the three Cmax/MIC targets by interval.")
q24 <- cmax_by_interval |> filter(interval == "q24h")
stopifnot(q24$`tenth dose` > q24$`first dose`)
```

The regimens the paper recommends (Table 5: 1800 mg q72h at CCr 40-50 mL/min
and 1800 mg q48h at CCr 60-90 mL/min) meet both the trough target and the
MIC 8 mg/L efficacy target in the simulation:

```{r table5}
recommended <- data.frame(
  interval = c("q72h", "q72h", rep("q48h", 4)),
  crcl = c(40, 50, 60, 70, 80, 90),
  dose = 1800
)
rec_chk <- cmp_first |>
  filter(target %in% c("cmin", "mic8")) |>
  inner_join(recommended, by = c("interval", "crcl", "dose")) |>
  select(interval, crcl, target, pta, paper) |>
  tidyr::pivot_wider(names_from = target, values_from = c(pta, paper))
rec_chk |>
  dplyr::rename("Interval" = interval, "CCr (mL/min)" = crcl,
                "PTA Cmin < 4 (sim)" = pta_cmin,
                "PTA Cmax/MIC >= 8, MIC 8 (sim)" = pta_mic8,
                "Paper Cmin flag" = paper_cmin,
                "Paper MIC 8 flag" = paper_mic8) |>
  knitr::kable(digits = 3, caption = "Table 5 recommended regimens.")
stopifnot(all(rec_chk$pta_cmin >= 0.9), all(rec_chk$pta_mic8 >= 0.9),
          all(rec_chk$paper_cmin == 1), all(rec_chk$paper_mic8 == 1))
```

The gates assert an overall agreement of at least 90%, and at least 85% for
each target, under the first-dose reading. A mutation control checks that the
comparison is not vacuous: every concentration is halved, as a two-fold error
in volume or dose unit would do, and the agreement must then fall below 80%.

```{r table3-gates}
cmp_mut <- cmp_first |>
  select(interval, crcl, dose, target, paper) |>
  inner_join(
    pta_grid(mutate(pta_wide, cmax_first = cmax_first / 2, cmin = cmin / 2),
             "cmax_first"),
    by = c("interval", "crcl", "dose", "target")
  )
mutated_agreement <- mean(cmp_mut$sim == cmp_mut$paper)
mutated_agreement

first_agree <- agreement |> filter(reading == "Cmax after the first dose")
stopifnot(
  first_agree$agreement[first_agree$target == "all"] >= 0.90,
  all(first_agree$agreement[first_agree$target != "all"] >= 0.85),
  mutated_agreement < 0.80
)
```

With the concentrations halved, the agreement falls to
`r round(100 * mutated_agreement)`%, against
`r round(100 * first_agree$agreement[first_agree$target == "all"])`% for the
model as published.

# Assumptions and deviations

* **Log-normal approximation of a non-parametric distribution.** Pmetrics
  NPAG estimates a discrete joint distribution of support points, which the
  paper does not publish. Table 2 summarises it by mean, SD, median and CV%
  per parameter. The model uses the mean as the typical value (as the abstract
  and Section 2.2 do) and converts each CV% to a log-normal variance with
  `omega^2 = log(CV^2 + 1)`. The two random effects are independent because no
  correlation is reported. Because the typical value is the
  median of a log-normal, the model's medians (2.25 L/h, 18.0 L) sit above
  the published medians (2.19 L/h, 17.1 L) and its means sit above the
  published means. The shape of the discrete NPAG distribution is not
  captured, which is the expected source of the remaining disagreement in the
  Table 3 cells near the 90% threshold.
* **Residual error not reported.** Neither the Pmetrics assay-error polynomial
  (C0-C3) nor the gamma or lambda noise term is reported. Both residual terms
  are therefore `fixed(0)`, so simulations from this model carry no
  measurement error. The paper's Table 3 targets are defined on the simulated
  concentrations themselves, so this does not affect the replication above.
* **Reference creatinine clearance.** The final-model equation in Section 2.2
  is `CL = CLs x (CCr/52.9)`, where 52.9 mL/min is the Table 1 cohort mean.
  The same section's prose says CCr was normalized to 'the median value of the
  study population of 52.1 mL/min' (the Table 1 median). The printed equation
  is used. The two differ by 1.5% in clearance at any given CCr.
* **Table 2 'Var' column.** The printed variances (2.19 for CL, 17.1 for V)
  repeat the median column and are not the squared SDs (0.61 and 11.6). They
  are not used; the CV% column (which matches SD / mean) is used instead.
* **Table 3 transcription and Cmax timing.** Table 3 is printed as raster
  images. The maintainers transcribed its 1080 `+`/`-` cells by classifying
  the glyphs in the embedded images; every row yielded exactly 40 cells. The
  paper does not say which dose the simulated Cmax follows. The first-dose
  reading is used for the gates and the tenth-dose reading is shown for
  comparison. The simulation uses 200 patients per creatinine clearance and
  interval rather than the paper's 1000.
* **Not reproduced.** The Figure 1 diagnostics and visual predictive check
  need the individual data, which are not published. The fractional target
  attainment of Table 4 needs the EUCAST *P. aeruginosa* MIC distribution the
  authors used, which is not given in the paper.
* **Covariate range.** The model was fit to CCr of 10.9-94.9 mL/min in
  patients aged 70 years and over, with no patient on renal replacement
  therapy. Clearance is proportional to CCr with no non-renal intercept, so it
  approaches zero as CCr approaches zero; do not extrapolate far below the
  observed range.
