---
title: "Cefiderocol in pneumonia, BSI/sepsis and cUTI (Kawaguchi 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cefiderocol in pneumonia, BSI/sepsis and cUTI (Kawaguchi 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Kawaguchi 2021 updates the same group's earlier cefiderocol population PK
analysis (`Kawaguchi_2018_cefiderocol_clcr` and its siblings) with the two
phase 3 studies, APEKS-NP (pneumonia) and CREDIBLE-CR (pneumonia, bloodstream
infection/sepsis or complicated urinary tract infection caused by
carbapenem-resistant pathogens). The structure is unchanged -- three
compartments, intravenous infusion, first-order elimination -- but the
covariate model is new: clearance now follows Cockcroft-Gault creatinine
clearance only up to 150 mL/min, each infection site carries its own factor on
clearance, and serum albumin enters the central volume.

```{r model}
mod <- rxode2::rxode(readModelDb("Kawaguchi_2021_cefiderocol"))
cat(mod$reference)
```

* Article: <https://doi.org/10.1128/AAC.01437-20>
* Supplement (Tables S1-S3 including the final NONMEM control stream,
  Figures S1-S5): available from the article page.

# Population

```{r population}
pop <- mod$population
str(pop[c("n_subjects", "n_studies", "n_observations", "age_range",
          "weight_range", "sex_female_pct")])
```

The analysis pooled `r pop$n_observations` plasma concentrations from
`r pop$n_subjects` subjects in five studies (Table 1, Table S1): 91 subjects
without infection from two phase 1 studies (healthy volunteers and a
renal-impairment study), 238 patients with cUTI or acute uncomplicated
pyelonephritis from the phase 2 APEKS-cUTI study, 115 pneumonia patients from
APEKS-NP, and 72 patients from CREDIBLE-CR (31 pneumonia, 20 BSI/sepsis, 21
cUTI). Body weight ranged 25-156 kg and Cockcroft-Gault creatinine clearance
4-306 mL/min; the phase 3 patients were markedly hypoalbuminaemic (median 2.7-3.0
g/dL against 4.2 g/dL in the earlier studies). Hemodialysis patients were
excluded. About 40% of subjects were female.

# Source trace

Every value below comes from Table 2 of the article (final model column) or
the final NONMEM control stream printed as Table S2. The two agree everywhere.

| Quantity | Value | Source |
|---|---|---|
| Structure: 3-compartment IV, first-order elimination | ADVAN11 TRANS4 | Results; Table S2 |
| CL (uninfected, CRCL 83.0 mL/min) | 4.04 L/h | Table 2; footnote b |
| V1 (uninfected, 72.6 kg, albumin 3.9 g/dL) | 7.78 L | Table 2; footnote b |
| Q2 | 6.19 L/h | Table 2 |
| V2 (72.6 kg) | 5.77 L | Table 2; footnote b |
| Q3 | 0.127 L/h | Table 2 |
| V3 | 0.798 L | Table 2 |
| CRCL exponent on CL, CRCL capped at 150 mL/min | 0.682 | Table 2; footnote b; Table S2 `CLCR1` |
| Weight exponent on V1 and V2 (shared) | 0.580 | Table 2; Table S2 `WT1` |
| Pneumonia factor on CL | 0.981 | Table 2; Table S2 `PT14` |
| BSI/sepsis factor on CL | 1.08 | Table 2; Table S2 `PT13` |
| cUTI (CREDIBLE-CR) factor on CL | 0.872 | Table 2; Table S2 `PT12` |
| cUTI/AUP (APEKS-cUTI) factor on CL | 1.27 | Table 2; Table S2 `PT11` |
| Albumin exponent on V1, centred at 3.9 g/dL | -0.617 | Table 2; footnote b; Table S2 `ALB2` |
| Any-infection factor on V1 | 1.39 | Table 2; Table S2 `PTV` |
| IIV CL, V1, V2 (CV%) | 37.5, 56.9, 33.6 | Table 2 |
| Covariances CL-V1, CL-V2, V1-V2 | 0.0886, 0.0792, 0.150 | Table 2 (footnotes c-e give R) |
| Proportional residual error | 20.5% | Table 2; Table S2 `W = IPRED*THETA(7)`, `$SIGMA 1 FIX` |
| Unbound fraction for fT>MIC | 0.422 | Materials and Methods |

## The omega scale

Table 2 prints IIV as a percent CV together with the three covariances and
their correlation coefficients. That over-determines the scale. If the variance
is `(CV/100)^2`, the implied covariance `R * sd_a * sd_b` must reproduce the
printed covariance; under the log-normal reading `sd = sqrt(log(1 + CV^2))` it
would be about 10% smaller.

```{r omega-scale}
cv <- c(CL = 0.375, V1 = 0.569, V2 = 0.336)
printed <- data.frame(
  pair = c("CL-V1", "CL-V2", "V1-V2"),
  a = c("CL", "CL", "V1"), b = c("V1", "V2", "V2"),
  cov = c(0.0886, 0.0792, 0.150),
  R = c(0.415, 0.629, 0.784)
)
sd_cv <- cv
sd_ln <- sqrt(log(1 + cv^2))
printed$cov_if_cv <- printed$R * sd_cv[printed$a] * sd_cv[printed$b]
printed$cov_if_lognormal <- printed$R * sd_ln[printed$a] * sd_ln[printed$b]
knitr::kable(printed[, c("pair", "cov", "cov_if_cv", "cov_if_lognormal")],
             digits = 4)
stopifnot(
  all(abs(printed$cov_if_cv / printed$cov - 1) < 0.005),
  all(abs(printed$cov_if_lognormal / printed$cov - 1) > 0.05)
)

# And the packaged omega matrix carries exactly those values.
om <- mod$omega
stopifnot(
  all(abs(diag(om) - cv^2) < 1e-12),
  abs(om["etalcl", "etalvc"] - 0.0886) < 1e-12,
  abs(om["etalcl", "etalvp"] - 0.0792) < 1e-12,
  abs(om["etalvc", "etalvp"] - 0.150) < 1e-12
)
```

# Deterministic verification

```{r helpers}
m0 <- suppressMessages(rxode2::zeroRe(mod))

# rxode2 warns when a zero-omega model is solved for many subjects; that is
# intended here, so only that one warning is muffled.
solve_quiet <- function(model, events, ...) {
  withCallingHandlers(
    as.data.frame(rxode2::rxSolve(model, events, returnType = "data.frame", ...)),
    warning = function(w) {
      if (grepl("without 'omega'", conditionMessage(w))) {
        invokeRestart("muffleWarning")
      }
    }
  )
}

cov_defaults <- c(DIS_PNEUMONIA = 0, DIS_BACTEREMIA = 0, DIS_CUTI = 0,
                  STUDY_CEFIDEROCOL_PHASE2 = 0)
```

## The Table 2 footnote equations, reproduced exactly

Footnote b of Table 2 prints the full covariate equations. The packaged model
must reproduce them to machine precision over a grid that crosses the 150 mL/min
clearance cap, spans body weight and albumin, and visits every infection site.

```{r gate-equations}
sites <- data.frame(
  site = c("none", "pneumonia", "BSI/sepsis", "cUTI (CREDIBLE-CR)", "cUTI/AUP (APEKS-cUTI)"),
  DIS_PNEUMONIA = c(0, 1, 0, 0, 0),
  DIS_BACTEREMIA = c(0, 0, 1, 0, 0),
  DIS_CUTI = c(0, 0, 0, 1, 1),
  STUDY_CEFIDEROCOL_PHASE2 = c(0, 0, 0, 0, 1),
  f_cl = c(1, 0.981, 1.08, 0.872, 1.27),
  f_v1 = c(1, 1.39, 1.39, 1.39, 1.39)
)
grid <- tidyr::expand_grid(
  CRCL = c(10, 83, 149, 150, 151, 300),
  WT = c(40, 72.6, 120),
  ALB = c(15, 39, 50),          # g/L
  sites
)
grid$id <- seq_len(nrow(grid))
ev_grid <- grid |>
  select(id, CRCL, WT, ALB, DIS_PNEUMONIA, DIS_BACTEREMIA, DIS_CUTI,
         STUDY_CEFIDEROCOL_PHASE2) |>
  mutate(time = 1, amt = 0, evid = 0L, cmt = "central")
typ <- solve_quiet(m0, ev_grid) |>
  select(id, cl, vc, vp)
chk <- grid |>
  inner_join(typ, by = "id") |>
  mutate(
    cl_paper = 4.04 * (pmin(CRCL, 150) / 83.0)^0.682 * f_cl,
    v1_paper = 7.78 * (WT / 72.6)^0.580 * ((ALB / 10) / 3.9)^-0.617 * f_v1,
    v2_paper = 5.77 * (WT / 72.6)^0.580
  )
c(cl = max(abs(chk$cl / chk$cl_paper - 1)),
  v1 = max(abs(chk$vc / chk$v1_paper - 1)),
  v2 = max(abs(chk$vp / chk$v2_paper - 1)))
stopifnot(
  nrow(chk) == nrow(grid),
  all(abs(chk$cl / chk$cl_paper - 1) < 1e-10),
  all(abs(chk$vc / chk$v1_paper - 1) < 1e-10),
  all(abs(chk$vp / chk$v2_paper - 1) < 1e-10)
)
```

## The paper's quoted percentages

The Results and Discussion state that CL in APEKS-cUTI patients was "27% higher
than that in subjects without infection" and that V1 in infected patients was
"39% higher". Clearance must also be flat above 150 mL/min.

```{r gate-quoted}
at <- function(site_name, crcl = 83) {
  chk[chk$site == site_name & chk$CRCL == crcl & chk$WT == 72.6 & chk$ALB == 39, ]
}
ratio_cl_cuti <- at("cUTI/AUP (APEKS-cUTI)")$cl / at("none")$cl
ratio_v1_inf <- at("pneumonia")$vc / at("none")$vc
flat <- at("none", 300)$cl / at("none", 150)$cl
c(cl_cuti = ratio_cl_cuti, v1_infected = ratio_v1_inf, cl_300_vs_150 = flat)
stopifnot(
  abs(ratio_cl_cuti - 1.27) < 1e-10,
  abs(ratio_v1_inf - 1.39) < 1e-10,
  abs(flat - 1) < 1e-12,
  at("none", 151)$cl == at("none", 150)$cl,
  at("none", 149)$cl < at("none", 150)$cl
)
```

## The infusion is an infusion

A single 2 g dose infused over 3 h in a typical pneumonia patient must peak at
the end of the infusion, not at time zero.

```{r gate-infusion}
ev_single <- data.frame(
  id = 1, time = c(0, seq(0.25, 48, by = 0.25)),
  amt = c(2000, rep(0, 192)), dur = c(3, rep(NA, 192)),
  evid = c(1L, rep(0L, 192)), cmt = "central",
  CRCL = 83, WT = 72.6, ALB = 28, DIS_PNEUMONIA = 1, DIS_BACTEREMIA = 0,
  DIS_CUTI = 0, STUDY_CEFIDEROCOL_PHASE2 = 0
)
single <- solve_quiet(m0, ev_single)
tmax <- single$time[which.max(single$Cc)]
tmax
stopifnot(tmax == 3)
```

# Virtual cohort

The paper's Monte-Carlo simulation (Tables 3-4 and Figure 4 captions; Materials
and Methods) generated, for each infection site, patients in six renal-function
groups defined by Cockcroft-Gault CRCL: augmented (120 to 150 and above 150
mL/min, half each), normal (90 to <120), mild (60 to <90), moderate (30 to <60),
severe (15 to <30) and ESRD (5 to <15). Body weight was log-normal with mean
72.6 kg and CV 30%, albumin log-normal with mean 2.8 g/dL and CV 30%. The cUTI
simulations used the CREDIBLE-CR coefficient (`STUDY_CEFIDEROCOL_PHASE2 = 0`).

The paper simulated 1,000 patients per scenario; this vignette uses 200. Both
the covariates **and the random effects** are drawn with base R, which is
reproducible across R builds, and the model is then solved with its random
effects zeroed and the drawn etas supplied as data columns. This keeps every
number below identical across rxode2 versions.

```{r cohort}
set.seed(20210217)
n_per_arm <- 200
regimens <- data.frame(
  renal = c("Augmented", "Normal", "Mild", "Moderate", "Severe", "ESRD"),
  crcl_lo = c(120, 90, 60, 30, 15, 5),
  crcl_hi = c(150, 120, 90, 60, 30, 15),
  dose_mg = c(2000, 2000, 2000, 1500, 1000, 750),
  tau = c(6, 8, 8, 8, 8, 12)
)
infection_sites <- c("Pneumonia", "BSI/sepsis", "cUTI")
sd_log <- function(cv) sqrt(log(1 + cv^2))
chol_om <- chol(mod$omega)

arms <- list()
for (s in infection_sites) {
  for (g in seq_len(nrow(regimens))) {
    r <- regimens[g, ]
    crcl <- if (r$renal == "Augmented") {
      # Half 120 to <150, half above 150. CL is flat above 150 mL/min, so the
      # upper bound of 200 mL/min only affects the covariate column.
      c(runif(n_per_arm / 2, 120, 150), runif(n_per_arm / 2, 150, 200))
    } else {
      runif(n_per_arm, r$crcl_lo, r$crcl_hi)
    }
    eta <- matrix(rnorm(3 * n_per_arm), n_per_arm) %*% chol_om
    arms[[length(arms) + 1]] <- data.frame(
      site = s, renal = r$renal, dose_mg = r$dose_mg, tau = r$tau,
      CRCL = crcl,
      WT = rlnorm(n_per_arm, log(72.6) - sd_log(0.3)^2 / 2, sd_log(0.3)),
      ALB = 10 * rlnorm(n_per_arm, log(2.8) - sd_log(0.3)^2 / 2, sd_log(0.3)),
      DIS_PNEUMONIA = as.integer(s == "Pneumonia"),
      DIS_BACTEREMIA = as.integer(s == "BSI/sepsis"),
      DIS_CUTI = as.integer(s == "cUTI"),
      STUDY_CEFIDEROCOL_PHASE2 = 0L,
      etalcl = eta[, 1], etalvc = eta[, 2], etalvp = eta[, 3]
    )
  }
}
cohort <- dplyr::bind_rows(arms)
cohort$id <- seq_len(nrow(cohort))
cohort$treatment <- paste(cohort$site, cohort$renal, sep = ": ")

cohort |>
  group_by(site) |>
  summarise(n = n(), WT_mean = mean(WT), ALB_mean_gdL = mean(ALB) / 10,
            .groups = "drop") |>
  knitr::kable(digits = 2)
stopifnot(
  nrow(cohort) == 18 * n_per_arm,
  !anyDuplicated(cohort$id),
  abs(mean(cohort$WT) / 72.6 - 1) < 0.05,
  abs(mean(cohort$ALB) / 28 - 1) < 0.05
)
```

# Simulation

Each subject is simulated over one steady-state dosing interval (3 h infusion,
`ss = 1`), observed every 0.05 h.

```{r simulate}
cov_cols <- c("CRCL", "WT", "ALB", "DIS_PNEUMONIA", "DIS_BACTEREMIA", "DIS_CUTI",
              "STUDY_CEFIDEROCOL_PHASE2", "etalcl", "etalvc", "etalvp")
dose_rows <- cohort |>
  transmute(id, time = 0, amt = dose_mg, dur = 3, ii = tau, ss = 1L,
            evid = 1L, cmt = "central") |>
  bind_cols(cohort[, cov_cols])
obs_rows <- cohort |>
  select(id, tau, all_of(cov_cols)) |>
  reframe(time = seq(0, tau, by = 0.05), .by = c(id, tau, all_of(cov_cols))) |>
  mutate(amt = 0, dur = NA_real_, ii = 0, ss = 0L, evid = 0L, cmt = "central") |>
  select(-tau)
events <- bind_rows(dose_rows, obs_rows) |>
  arrange(id, time, desc(evid))

# maxsteps: the lsoda step budget is charged across the steady-state search.
sim <- solve_quiet(m0, events, maxsteps = 1e6) |>
  select(id, time, Cc, cl) |>
  inner_join(cohort |> select(id, site, renal, treatment, tau, dose_mg), by = "id")
stopifnot(!anyNA(sim$Cc), all(sim$Cc > 0))
```

## Typical profiles by infection site

```{r figure-profiles, fig.height = 4}
typ_ev <- sites |>
  filter(site != "cUTI/AUP (APEKS-cUTI)") |>
  mutate(id = row_number(), CRCL = 83, WT = 72.6, ALB = 28)
typ_ev <- typ_ev |>
  select(-f_cl, -f_v1) |>
  tidyr::crossing(time = seq(0, 48, by = 0.1)) |>
  mutate(amt = 0, dur = NA_real_, ii = 0, addl = 0L, evid = 0L, cmt = "central") |>
  bind_rows(
    typ_ev |>
      select(-f_cl, -f_v1) |>
      mutate(time = 0, amt = 2000, dur = 3, ii = 8, addl = 5L, evid = 1L,
             cmt = "central")
  ) |>
  arrange(id, time, desc(evid))
typ_sim <- solve_quiet(m0, typ_ev, keep = "site")
ggplot(typ_sim, aes(time, Cc, colour = site)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Total cefiderocol (ug/mL)", colour = NULL,
       caption = paste("Typical patient, CRCL 83 mL/min, 72.6 kg, albumin",
                       "2.8 g/dL, 2 g q8h over 3 h.")) +
  theme_bw()
```

# Replicate published tables

## Tables 3 and 4 -- probability of target attainment

The free concentration is `0.422 * Cc` (Materials and Methods). The percentage
of the dosing interval with free concentration above the MIC (%fT>MIC) is the
fraction of the 0.05 h grid over `[0, tau)` above the MIC; 100% fT>MIC is
equivalent to a free trough above the MIC.

```{r pta}
mics <- c(0.25, 0.5, 1, 2, 4, 8, 16)
ft <- sim |>
  filter(time < tau) |>
  mutate(Cfree = 0.422 * Cc) |>
  tidyr::crossing(mic = mics) |>
  group_by(id, site, renal, mic) |>
  summarise(ft = mean(Cfree > mic), .groups = "drop")
pta_sim <- ft |>
  group_by(site, renal, mic) |>
  summarise(pta75 = 100 * mean(ft >= 0.75), pta100 = 100 * mean(ft >= 1),
            .groups = "drop")

# Kawaguchi 2021 Table 3 (75% fT>MIC) and Table 4 (100% fT>MIC).
pta_row <- function(site, renal, target, v) {
  data.frame(site = site, renal = renal, target = target, mic = mics, paper = v)
}
paper <- rbind(
  pta_row("Pneumonia", "Augmented", 75, c(100, 100, 100, 100, 99.7, 94.5, 60.4)),
  pta_row("Pneumonia", "Normal", 75, c(100, 100, 100, 99.9, 98.9, 87.1, 43.4)),
  pta_row("Pneumonia", "Mild", 75, c(100, 100, 100, 100, 99.8, 97.0, 69.7)),
  pta_row("Pneumonia", "Moderate", 75, c(100, 100, 100, 100, 99.9, 98.7, 83.3)),
  pta_row("Pneumonia", "Severe", 75, c(100, 100, 100, 100, 100, 99.9, 90.7)),
  pta_row("Pneumonia", "ESRD", 75, c(100, 100, 100, 100, 100, 99.6, 86.3)),
  pta_row("BSI/sepsis", "Augmented", 75, c(100, 100, 100, 100, 99.4, 91.3, 49.6)),
  pta_row("BSI/sepsis", "Normal", 75, c(100, 100, 100, 99.9, 97.3, 80.6, 32.6)),
  pta_row("BSI/sepsis", "Mild", 75, c(100, 100, 100, 99.9, 99.6, 94.4, 57.7)),
  pta_row("BSI/sepsis", "Moderate", 75, c(100, 100, 100, 100, 99.9, 98.0, 74.8)),
  pta_row("BSI/sepsis", "Severe", 75, c(100, 100, 100, 100, 100, 99.8, 84.8)),
  pta_row("BSI/sepsis", "ESRD", 75, c(100, 100, 100, 100, 100, 99.2, 79.2)),
  pta_row("cUTI", "Augmented", 75, c(100, 100, 100, 100, 99.9, 96.9, 73.3)),
  pta_row("cUTI", "Normal", 75, c(100, 100, 100, 100, 99.6, 93.6, 56.3)),
  pta_row("cUTI", "Mild", 75, c(100, 100, 100, 100, 99.8, 98.4, 81.2)),
  pta_row("cUTI", "Moderate", 75, c(100, 100, 100, 100, 100, 99.6, 90.4)),
  pta_row("cUTI", "Severe", 75, c(100, 100, 100, 100, 100, 100, 95.9)),
  pta_row("cUTI", "ESRD", 75, c(100, 100, 100, 100, 100, 100, 91.6)),
  pta_row("Pneumonia", "Augmented", 100, c(100, 100, 100, 99.7, 95.9, 79.8, 37.0)),
  pta_row("Pneumonia", "Normal", 100, c(100, 100, 99.9, 98.3, 91.2, 64.6, 23.2)),
  pta_row("Pneumonia", "Mild", 100, c(100, 100, 99.9, 99.7, 98.2, 85.9, 46.4)),
  pta_row("Pneumonia", "Moderate", 100, c(100, 100, 100, 100, 99.5, 94.8, 66.7)),
  pta_row("Pneumonia", "Severe", 100, c(100, 100, 100, 100, 100, 99.5, 81.8)),
  pta_row("Pneumonia", "ESRD", 100, c(100, 100, 100, 100, 100, 98.3, 77.1)),
  pta_row("BSI/sepsis", "Augmented", 100, c(100, 100, 100, 99.4, 93.6, 71.6, 28.5)),
  pta_row("BSI/sepsis", "Normal", 100, c(100, 99.9, 99.5, 96.2, 85.8, 54.0, 14.1)),
  pta_row("BSI/sepsis", "Mild", 100, c(100, 100, 99.8, 99.4, 96.0, 78.0, 36.1)),
  pta_row("BSI/sepsis", "Moderate", 100, c(100, 100, 100, 99.9, 98.7, 91.2, 55.8)),
  pta_row("BSI/sepsis", "Severe", 100, c(100, 100, 100, 100, 100, 98.3, 74.7)),
  pta_row("BSI/sepsis", "ESRD", 100, c(100, 100, 100, 100, 100, 96.8, 68.0)),
  pta_row("cUTI", "Augmented", 100, c(100, 100, 100, 100, 98.0, 88.3, 51.1)),
  pta_row("cUTI", "Normal", 100, c(100, 100, 99.9, 99.4, 95.1, 77.6, 34.3)),
  pta_row("cUTI", "Mild", 100, c(100, 100, 100, 99.8, 98.9, 93.2, 59.4)),
  pta_row("cUTI", "Moderate", 100, c(100, 100, 100, 100, 99.8, 97.7, 79.1)),
  pta_row("cUTI", "Severe", 100, c(100, 100, 100, 100, 100, 99.7, 90.1)),
  pta_row("cUTI", "ESRD", 100, c(100, 100, 100, 100, 100, 99.4, 85.7))
)
stopifnot(nrow(paper) == 2 * 18 * length(mics))

pta_cmp <- pta_sim |>
  tidyr::pivot_longer(c(pta75, pta100), names_to = "target", values_to = "sim") |>
  mutate(target = ifelse(target == "pta75", 75, 100)) |>
  inner_join(paper, by = c("site", "renal", "target", "mic")) |>
  mutate(diff = sim - paper)
stopifnot(nrow(pta_cmp) == nrow(paper))
```

```{r pta-table}
pta_cmp |>
  filter(mic %in% c(4, 8, 16)) |>
  mutate(cell = sprintf("%.1f (%.1f)", sim, paper)) |>
  select(target, site, renal, mic, cell) |>
  tidyr::pivot_wider(names_from = mic, values_from = cell) |>
  dplyr::rename(
    "Target %fT>MIC" = target, "Infection site" = site,
    "Renal function" = renal, "MIC 4" = `4`, "MIC 8" = `8`, "MIC 16" = `16`
  ) |>
  knitr::kable(caption = paste(
    "Simulated PTA (%) with Kawaguchi 2021 Tables 3 and 4 in parentheses,",
    "for the MICs where attainment is not saturated at 100%.",
    "200 simulated patients per row against 1,000 in the paper."
  ))
```

```{r figure-pta, fig.height = 7}
ggplot(pta_cmp, aes(mic)) +
  geom_line(aes(y = sim, colour = "Simulated")) +
  geom_point(aes(y = paper, colour = "Kawaguchi 2021")) +
  facet_grid(renal ~ paste0(target, "% fT>MIC, ", site)) +
  scale_x_log10(breaks = mics, labels = format(mics)) +
  labs(x = "MIC (ug/mL)", y = "PTA (%)", colour = NULL,
       caption = "Replicates Tables 3 and 4 of Kawaguchi 2021.") +
  theme_bw() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))
```

The agreement is measured over every cell, and separately over the cells whose
paper value is below 99% (where attainment is not saturated and the comparison
actually discriminates). With 200 patients per row the binomial standard error
at a 50% PTA is 3.5 percentage points, so a few points of scatter is expected.

```{r gate-pta}
disc <- pta_cmp |> filter(paper < 99)
summary_pta <- c(
  n_cells = nrow(pta_cmp),
  n_discriminating = nrow(disc),
  median_diff = median(disc$diff),
  median_abs_diff = median(abs(disc$diff)),
  p90_abs_diff = unname(quantile(abs(disc$diff), 0.9)),
  max_abs_diff_all = max(abs(pta_cmp$diff))
)
round(summary_pta, 2)
stopifnot(
  nrow(disc) >= 30,
  # Centre: a mis-transcribed clearance or unbound fraction shifts every
  # non-saturated cell in one direction by far more than this.
  abs(median(disc$diff)) < 3,
  median(abs(disc$diff)) < 4,
  quantile(abs(disc$diff), 0.9) < 8
)
```

## Figure 4 -- integrated PTA and the text's breakpoint claims

The integrated PTA weights the renal-function groups by the phase 3 CRCL
distribution (Materials and Methods: 20.3%, 15.0%, 24.6%, 32.6%, 4.8% and 2.7%
from augmented to ESRD). The Results state that the highest MIC reaching >90%
integrated PTA was 8 ug/mL for 75% fT>MIC regardless of infection site, and 4
ug/mL for 100% fT>MIC. Figure 4 itself is not digitised here; instead the same
weighting is applied both to the simulation and to the paper's Tables 3-4.

```{r figure-4}
weights <- data.frame(
  renal = regimens$renal,
  w = c(20.3, 15.0, 24.6, 32.6, 4.8, 2.7) / 100
)
integ <- pta_cmp |>
  inner_join(weights, by = "renal") |>
  group_by(site, target, mic) |>
  summarise(sim = sum(w * sim) / sum(w), paper = sum(w * paper) / sum(w),
            .groups = "drop")
ggplot(integ, aes(mic, sim, colour = site)) +
  geom_line() +
  geom_point(aes(y = paper), shape = 1) +
  geom_hline(yintercept = 90, linetype = 2) +
  facet_wrap(~ paste0(target, "% fT>MIC")) +
  scale_x_log10(breaks = mics, labels = format(mics)) +
  labs(x = "MIC (ug/mL)", y = "Integrated PTA (%)", colour = NULL,
       caption = paste("Lines: simulated. Open circles: the same weighting",
                       "applied to Tables 3-4. Replicates Figure 4.")) +
  theme_bw()

highest_mic <- integ |>
  group_by(site, target) |>
  summarise(
    simulated = max(mic[sim > 90]),
    from_tables = max(mic[paper > 90]),
    sim_at_8 = sim[mic == 8], tables_at_8 = paper[mic == 8],
    .groups = "drop"
  )
knitr::kable(highest_mic, digits = 1, caption = paste(
  "Highest MIC with integrated PTA above 90%, from the simulation and from",
  "the same weighting applied to Tables 3-4, with the integrated PTA at",
  "MIC 8 ug/mL."
))
stopifnot(
  nrow(highest_mic) == 6,
  # Integrated PTA tracks the paper's tables at every MIC.
  max(abs(integ$sim - integ$paper)) < 4,
  # 75% fT>MIC: 8 ug/mL for every infection site, as the Results state.
  all(highest_mic$simulated[highest_mic$target == 75] == 8),
  # 100% fT>MIC: 4 ug/mL for pneumonia and BSI/sepsis.
  all(highest_mic$simulated[highest_mic$target == 100 & highest_mic$site != "cUTI"] == 4)
)
```

For 100% fT>MIC in cUTI the text's "4 ug/mL" cannot be reproduced from the
paper's own Table 4: weighting its cUTI rows gives
`r round(highest_mic$tables_at_8[highest_mic$site == "cUTI" & highest_mic$target == 100], 1)`%
at 8 ug/mL, just above the 90% line, and the simulation gives
`r round(highest_mic$sim_at_8[highest_mic$site == "cUTI" & highest_mic$target == 100], 1)`%.
The cUTI cell sits on the 90% threshold, so its "highest MIC" is not a stable
quantity and is not asserted; its integrated PTA is covered by the
`max(abs(sim - paper))` check above.

# PKNCA validation

NCA over the steady-state dosing interval for the pneumonia arms (the largest
infected population in the analysis).

```{r pknca}
nca_ids <- cohort |> filter(site == "Pneumonia") |> pull(id)
sim_nca <- sim |>
  filter(id %in% nca_ids, abs(time / 0.25 - round(time / 0.25)) < 1e-8) |>
  select(id, time, Cc, treatment)
dose_df <- cohort |>
  filter(id %in% nca_ids) |>
  transmute(id, time = 0, amt = dose_mg, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "ug/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "mg")
intervals <- cohort |>
  filter(id %in% nca_ids) |>
  distinct(treatment, tau) |>
  transmute(treatment, start = 0, end = tau, cmax = TRUE, tmax = TRUE,
            cmin = TRUE, cav = TRUE, auclast = TRUE)
nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)
nca_tbl <- as.data.frame(nca_res$result)
stopifnot(nrow(nca_tbl) > 0, !anyNA(nca_tbl$PPORRES))
```

## Steady-state mass balance

At steady state, `CL * AUC(0-tau) = Dose` for every subject.

```{r gate-mass-balance}
cl_ind <- sim |> filter(id %in% nca_ids) |> distinct(id, cl, dose_mg)
mb <- nca_tbl |>
  filter(PPTESTCD == "auclast") |>
  select(id, auc = PPORRES) |>
  inner_join(cl_ind, by = "id") |>
  mutate(pct = 100 * (cl * auc / dose_mg - 1))
summary(mb$pct)
stopifnot(
  nrow(mb) == length(nca_ids),
  abs(median(mb$pct)) < 1,
  quantile(abs(mb$pct), 0.95) < 2
)
```

## Comparison against the published exposure

The Discussion reports geometric-mean daily AUC from the phase 3 post hoc
estimates: 1,365 `ug*h/mL` at 2 g q6h in patients with augmented renal function,
and 1,494 `ug*h/mL` at 2 g q8h in patients with normal renal function. These are
empirical summaries of the enrolled patients (whose covariates are not
published), not typical-value predictions, so the comparison is corroborative.
Daily AUC is `24 * Cavg`; the published values are shown as the equivalent
Cavg. The 2 g q8h group is taken as the simulated normal and mild arms together
(CRCL 60 to <120 mL/min, the range that receives 2 g q8h).

```{r nca-compare}
simulated <- nca_tbl |>
  filter(PPTESTCD == "cav") |>
  mutate(group = case_when(
    grepl("Augmented", treatment) ~ "2 g q6h, augmented",
    grepl("Normal|Mild", treatment) ~ "2 g q8h, normal",
    TRUE ~ NA_character_
  )) |>
  filter(!is.na(group)) |>
  group_by(group, PPTESTCD) |>
  summarise(PPORRES = exp(mean(log(PPORRES))), .groups = "drop")
published <- data.frame(
  group = c("2 g q6h, augmented", "2 g q8h, normal"),
  cav = c(1365, 1494) / 24
)
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = simulated,
  reference = published,
  by = "group",
  units = c(cav = "ug/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = paste(
  "Simulated geometric-mean steady-state Cavg (pneumonia patients) against",
  "the Discussion's phase 3 geometric-mean daily AUC divided by 24 h.",
  "* differs from reference by more than 20%."
))
sim_cav <- setNames(simulated$PPORRES, simulated$group)
stopifnot(all(abs(sim_cav[published$group] / published$cav - 1) < 0.2))
```

# Assumptions and deviations

* **Covariate encoding.** The paper's infection-site variable (`PT` in Table S2)
  is split into mutually exclusive indicators `DIS_PNEUMONIA`, `DIS_BACTEREMIA`
  (the paper's BSI/sepsis category) and `DIS_CUTI` (cUTI, pooled with acute
  uncomplicated pyelonephritis in the phase 2 study). The any-infection factor
  on V1 is applied to their sum. The pneumonia arm pooled hospital-acquired,
  ventilator-associated and health care-associated pneumonia under one
  coefficient, so the general pneumonia flag is used.
* **Study-specific cUTI effect.** The cUTI factor on CL differs between the
  phase 2 APEKS-cUTI study (1.27) and the phase 3 CREDIBLE-CR study (0.872); the
  control stream switches on the subject ID. The packaged model uses a
  `STUDY_CEFIDEROCOL_PHASE2` indicator, whose default of 0 gives the CREDIBLE-CR
  value that the paper used in its own simulations.
* **Albumin units.** The model was calibrated in g/dL (reference 3.9 g/dL). The
  packaged model takes albumin in g/L and divides by 10 internally.
* **Albumin exponent sign.** The article's typeset minus sign (-0.617) is
  confirmed by the control stream's bounds for THETA(14), (-2.0, -0.5, 0.5),
  and by the Discussion (lower albumin, larger V1).
* **Time-varying creatinine clearance.** The source data used baseline CRCL for
  the phase 1 and phase 2 studies and time-varying CRCL for the phase 3 studies;
  the packaged model accepts either.
* **Virtual cohort details not stated in the paper.** The log-normal body
  weight and albumin distributions are parameterised with the stated means as
  arithmetic means. The upper bound for the augmented group above 150 mL/min is
  set at 200 mL/min, which has no effect on CL because of the 150 mL/min cap.
  CRCL is uniform within each group. The paper's %fT>MIC is taken without
  residual error; simulating 200 rather than 1,000 patients per scenario adds a
  few points of binomial scatter to each PTA cell.
* **Integrated PTA claim for cUTI.** The Results state that the highest MIC
  with more than 90% integrated PTA for 100% fT>MIC was 4 ug/mL. Applying the
  stated renal-function weights to the paper's own Table 4 gives 91.8% for cUTI
  at 8 ug/mL, so for cUTI the statement holds only if Figure 4 was computed
  slightly differently from the tables. Pneumonia and BSI/sepsis agree with the
  statement, and the simulated cUTI value (about 91%) sits on the same
  threshold as the tables.
