---
title: "Dolutegravir (Kawuma 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dolutegravir (Kawuma 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Kawuma_2023_dolutegravir"))$reference`
* Description: `r rxode2::rxode(readModelDb("Kawuma_2023_dolutegravir"))$description`
* Article: <https://doi.org/10.1111/bcp.15604>
* Predecessor model (source of the fixed allometric exponents, the random-effect
  structure, the residual-error form and the assay LLOQ values):
  <https://doi.org/10.1128/aac.00215-22>

Dolutegravir is the WHO-recommended anchor drug for first- and second-line
antiretroviral therapy, and tuberculosis is the commonest opportunistic
infection among people living with HIV. Rifampicin, the cornerstone of
tuberculosis treatment, is a potent inducer of the UGT1A1 and CYP3A4 pathways
that clear dolutegravir, so co-treated patients need twice-daily dolutegravir.
Rifabutin is a less potent inducer and is a candidate substitute. This paper
extends the authors' earlier dolutegravir-rifampicin population PK model with
the dolutegravir concentrations measured during rifabutin co-administration in
arm B of NCT01231542, which the earlier analysis had excluded.

The headline result is mechanistically unusual: rifabutin's effect is best
described not on clearance but on the **apparent central volume of
distribution**, which falls by 33.1%. Because clearance is untouched, steady
state AUC is unchanged while the concentration-time profile becomes peakier --
higher Cmax, lower Cmin. The gates below are built around exactly that
signature.

# Population

The model was fit to 907 dolutegravir plasma concentrations from 41 healthy
HIV-negative volunteers (68% male) enrolled in two studies (Kawuma 2023
Results); 90 of those samples were taken during rifabutin co-administration.
Median (interquartile range) weight was 81.5 (69.5-88.6) kg and median age was
43 (31-50) years. No sample was below the lower limit of quantification.

* **RADIO** (n = 16) was a phase II open-label sequential healthy-volunteer
  study run at the St Stephen's Centre, London. Volunteers received dolutegravir
  50 mg once daily for 7 days, 100 mg once daily for 7 days, rifampicin 600 mg
  once daily alone for 14 days, then dolutegravir 50 mg plus rifampicin for
  7 days and finally dolutegravir 100 mg plus rifampicin for 7 days. Doses were
  taken after a standard breakfast.
* **NCT01231542** (n = 25) was a phase I open-label two-arm fixed-sequence
  study run in Baltimore. Arm A (n = 12) received dolutegravir 50 mg once daily,
  then 50 mg twice daily, then 50 mg twice daily with rifampicin 600 mg once
  daily. Arm B (n = 13) received dolutegravir 50 mg once daily, then 50 mg once
  daily with rifabutin 300 mg once daily for 14 days. Doses were taken after an
  overnight fast.

Intensive pharmacokinetic sampling was done on the last day of each regimen.
The same information is available programmatically via
`readModelDb("Kawuma_2023_dolutegravir")()$population`.

```{r population}
pop <- rxode2::rxode(readModelDb("Kawuma_2023_dolutegravir"))$population
str(pop)
```

# Source trace

Every `ini()` value carries an in-file comment naming its source location in
`inst/modeldb/specificDrugs/Kawuma_2023_dolutegravir.R`. They are collected
here for review. "Table 1" means Table 1 of Kawuma 2023; the two rows sourced
from the predecessor paper are marked.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` | `log(1.03)` | Table 1, `CL/F (L/h)` = 1.03 (0.945, 1.15) |
| `lvc` | `log(13.3)` | Table 1, `Vc/F (L)` = 13.3 (11.7, 14.5) |
| `lq` | `log(0.675)` | Table 1, `Q/F (L/h)` = 0.675 (0.439, 1.02) |
| `lvp` | `log(3.52)` | Table 1, `Vp/F (L)` = 3.52 (2.69, 4.24) |
| `lka` | `log(1.63)` | Table 1, `Absorption rate constant (/h)` = 1.63 (1.16, 2.73) |
| `lfdepot` | `fixed(log(1))` | Table 1, `Relative bioavailability` = 1 fixed |
| `ltlag` | `log(0.205)` | Table 1, `Absorption lag time NCT01231542 study (h)` = 0.205 (0.00205, 0.381), footnote c "fasted conditions" |
| `e_fed_tlag` | `log(0.986 / 0.205)` | Table 1, `Absorption lag time RADIO study (h)` = 0.986 (0.602, 1.43), footnote d "Under fed condition" |
| `e_wt_cl` | `fixed(0.75)` | Table 1 footnote b (allometric scaling, 70 kg reference); exponent value from **Kawuma 2022** Methods, "allometric exponents for clearance and volume parameters were fixed at 0.75 and 1" |
| `e_wt_vc` | `fixed(1)` | as above (**Kawuma 2022** Methods) |
| `e_rifampicin_cl` | `1.43` | Table 1, `Rifampicin co-administration on CL (%)` = +143 (+126, +155) |
| `e_rifabutin_vc` | `-0.331` | Table 1, `Rifabutin co-administration on Vc (%)` = -33.1 (-25.1, -42.3); also Results text |
| `e_sexf_ka` | `-0.381` | Table 1, `Male sex on ka (%)` = -38.1 (-15.1, -56.1) |
| `etalcl` | `0.063001` | Table 1, `BSV clearance` = 25.1 (15.7, 34.4) %CV; footnote e gives CV% = sqrt(omega^2) x 100, so omega^2 = 0.251^2 |
| `etaiov_fdepot_*` | `0.1849` | Table 1, `BOV bioavailability` = 43.0 (37.0, 49.6) %CV; omega^2 = 0.430^2 |
| `etaiov_tlag_*` | `0.219961` | Table 1, `BOV lag time` = 46.9 (16.5, 79.8) %CV; omega^2 = 0.469^2 |
| `etaiov_ka_*` | `0.368449` | Table 1, `BOV absorption rate constant` = 60.7 (41.9, 79.2) %CV; omega^2 = 0.607^2 |
| `propSd` | `0.0885` | Table 1, `proportional error (%)` = 8.85 (7.32, 9.79) |
| `addSd` | `0.036` | Table 1, `Additive error NCT01231542 study (mg/L)` = 0.036 (0.00435, 0.0801) |
| `addSd_radio` | `0.0485` | Table 1, `Additive error RADIO study (mg/L)` = 0.0485 (0.0103, 0.0861) |
| Two-compartment disposition, first-order elimination, lagged first-order absorption | n/a | Kawuma 2023 Results, "The model consists of two compartments with first-order elimination and lagged first-order absorption" |
| Log-normal BSV on disposition, BOV on absorption parameters | n/a | **Kawuma 2022** Methods, "BOV was incorporated on absorption parameters, while BSV was tested on disposition parameters" |
| Combined additive + proportional residual error, additive floored at 0.2 x LLOQ | n/a | **Kawuma 2022** Methods and "Analytical assay" (LLOQ 0.020 mg/L for NCT01231542, 0.050 mg/L for RADIO) |

The two `Kawuma 2022` rows are the only structural facts not printed in the
2023 paper itself; every parameter **value** comes from the 2023 Table 1.

```{r model-load}
mod <- readModelDb("Kawuma_2023_dolutegravir")
ui <- rxode2::rxode(mod)
mod_typical <- rxode2::zeroRe(ui)
```

# Deterministic replication of the rifabutin interaction

The paper's central quantitative claims are relative changes in steady-state
exposure caused by rifabutin (Kawuma 2023 Results):

> As a consequence of this effect on volume, the model predicts that rifabutin
> co-administration increases dolutegravir's Cmax by 15.1% (95% CI: -49.8 to
> 268), reduces Cmin by 30.1% (95% CI: 24.7-51.5) and has no effect on the area
> under the concentration-time curve (AUC).

Those point estimates are medians over a simulated cohort, so they are checked
against a cohort below. First, the deterministic (typical-value) prediction --
which has no Monte-Carlo noise at all and is therefore the sharper gate on
structure, units and dosing.

```{r typical-ddi}
# Steady state under 50 mg once daily, 70 kg, fasted, no rifampicin. The run-in
# is 10 days: the slowest plausible subject (BSV CL at -2 SD, CL/F = 0.62 L/h)
# has a terminal half-life near 19 h, so 240 h is >12 half-lives even in the
# tail of the population.
runin_days <- 10
t0 <- runin_days * 24

obs_grid <- c(seq(0, 6, by = 0.05), seq(6.5, 24, by = 0.5))

typical_profile <- function(dose, ii, rifabutin = 0, rifampicin = 0,
                            sexf = 0, wt = 70, fed = 0) {
  ev <- rxode2::et(amt = dose, ii = ii, until = t0 + ii, cmt = "depot") |>
    rxode2::et(t0 + obs_grid)
  d <- as.data.frame(ev)
  d$WT <- wt
  d$SEXF <- sexf
  d$FED <- fed
  d$STUDY_RADIO <- 0
  d$OCC <- 1
  d$CONMED_RIFAMPICIN <- rifampicin
  d$CONMED_RIFABUTIN <- rifabutin
  rxode2::rxSolve(
    mod_typical, d, returnType = "data.frame", maxsteps = 200000L,
    atol = 1e-10, rtol = 1e-8
  ) |>
    dplyr::filter(time >= t0) |>
    dplyr::mutate(time = time - t0)
}

# Trapezoidal AUC over the dosing interval of an already-steady-state profile.
auc_tau <- function(df, tau) {
  d <- df |> dplyr::filter(time <= tau) |> dplyr::arrange(time)
  sum(diff(d$time) * (utils::head(d$Cc, -1) + utils::tail(d$Cc, -1)) / 2)
}

alone <- typical_profile(50, 24, rifabutin = 0)
withrfb <- typical_profile(50, 24, rifabutin = 1)

typical_ddi <- tibble::tibble(
  Metric = c("Cmax (mg/L)", "Cmin (mg/L)", "AUC0-24 (mg*h/L)"),
  `DTG alone` = c(max(alone$Cc), min(alone$Cc), auc_tau(alone, 24)),
  `DTG + rifabutin` = c(max(withrfb$Cc), min(withrfb$Cc), auc_tau(withrfb, 24))
) |>
  dplyr::mutate(
    `Change (%)` = 100 * (`DTG + rifabutin` / `DTG alone` - 1),
    `Kawuma 2023 (%)` = c(15.1, -30.1, 0)
  )

knitr::kable(typical_ddi, digits = c(0, 3, 3, 1, 1),
             caption = "Typical-value steady-state exposure, dolutegravir 50 mg once daily in a fasted 70 kg man, with and without rifabutin 300 mg once daily. The last column is the change reported in Kawuma 2023 Results.")
```

Two things to read off that table.

**AUC is untouched, to four decimal places.** This is the sharpest structural
check in the vignette. At steady state the AUC over one dosing interval is
`F * Dose / (CL/F)`, so a model in which rifabutin acts *only* on `Vc` must
leave it exactly unchanged. Had the effect been transcribed onto clearance --
the alternative parameterisation the authors explicitly fitted and rejected --
AUC would have moved by tens of percent. The same closed form also pins the
absolute AUC against the printed clearance of 1.03 L/h, independently of the
ODE solver, the depot, the lag time and the peripheral compartment.

```{r closed-form-auc}
# AUC0-tau at steady state = F * Dose / CL, with F fixed at 1 and CL/F = 1.03
# L/h for a typical 70 kg individual (Kawuma 2023 Table 1). Nothing here is
# read back out of the solve -- the expected value is arithmetic on the
# printed parameter.
auc_closed_form <- 50 / 1.03

closed_form <- tibble::tibble(
  Arm = c("DTG alone", "DTG + rifabutin"),
  `Solved AUC0-24` = c(auc_tau(alone, 24), auc_tau(withrfb, 24)),
  `Dose / CL` = auc_closed_form
) |>
  dplyr::mutate(`Difference (%)` = 100 * (`Solved AUC0-24` / `Dose / CL` - 1))

knitr::kable(closed_form, digits = 4,
             caption = "Steady-state AUC against its closed form. Deterministic; the residual is pure trapezoidal-integration error.")

stopifnot(
  # Deterministic quantities: no Monte-Carlo noise, so these are tight. The
  # realised residual is ~0.005%, i.e. trapezoidal error on the 0.05 h grid.
  all(abs(closed_form$`Difference (%)`) < 0.2),
  # Rifabutin must not move AUC at all. A transcription of the effect onto
  # clearance instead of volume would move this by ~25-40%.
  abs(100 * (auc_tau(withrfb, 24) / auc_tau(alone, 24) - 1)) < 0.1
)
```

**Cmax rises and Cmin falls, by roughly the reported amounts.** The
typical-value changes are +22.8% and -26.1% against the paper's +15.1% and
-30.1%. These are not the same statistic -- the paper reports the median of the
per-subject ratio across a simulated cohort, and the between-occasion
variability on all three absorption parameters makes that median differ
appreciably from the typical-value prediction (the paper's own 95% interval on
the Cmax change runs from -49.8% to +268%). The cohort comparison below is the
like-for-like one.

```{r typical-ddi-gates}
cmax_change <- 100 * (max(withrfb$Cc) / max(alone$Cc) - 1)
cmin_change <- 100 * (min(withrfb$Cc) / min(alone$Cc) - 1)

stopifnot(
  # Deterministic. Bands are wide enough to contain both the typical-value
  # prediction (+22.8 / -26.1) and the paper's cohort median (+15.1 / -30.1),
  # and narrow enough that a lost or inverted volume effect (0% change) or a
  # clearance-sited effect (Cmax and Cmin both falling ~40%) breaks them.
  cmax_change > 8 && cmax_change < 35,
  cmin_change < -15 && cmin_change > -40
)
```

# Virtual cohort

Original observed data are not publicly available. The cohort below mirrors the
simulation the authors report: 70 kg individuals dosed with dolutegravir 50 mg
once daily with and without rifabutin (Kawuma 2023 Methods). Doses are fasted,
following the predecessor paper's simulation convention (Kawuma 2022 Figure 3
caption: "For the absorption lag time, we used the value estimated for the
NCT01231542 study since this was done under fasted conditions"). Sex is drawn
at the studies' 68% male / 32% female split; the paper does not state the sex
composition of its simulated cohort, and sex enters only through `ka`, which
barely moves a steady-state trough.

The two arms are a **paired** within-subject comparison, matching the
fixed-sequence design of NCT01231542 arm B, in which each volunteer took
dolutegravir alone and then dolutegravir with rifabutin. The two arms reuse the
same subject IDs and the same `rxSetSeed()`, so between-subject clearance is
shared, while `OCC = 1` and `OCC = 2` select independent between-occasion draws
for bioavailability, lag time and absorption rate -- exactly the structure the
model encodes.

```{r cohort}
n_subj <- 200

# set.seed() seeds R's RNG (used here only for the sex draw). It does NOT seed
# rxode2's simulation RNG, and rxode2's streams are partitioned per solver
# thread -- so the etas drawn below are reproducible on this machine and
# different on a machine with a different thread count. Every assertion in this
# vignette is written to hold for any cohort the model can produce; see
# pattern 12 of the extraction skill's known-vignette-failure-patterns.
set.seed(20260905)
sex_draw <- stats::rbinom(n_subj, 1, 0.32)

ddi_arm <- function(rifabutin, occ) {
  rxode2::rxSetSeed(4242)
  ev <- rxode2::et(amt = 50, ii = 24, until = t0 + 24, cmt = "depot") |>
    rxode2::et(t0 + obs_grid) |>
    rxode2::et(id = seq_len(n_subj))
  d <- as.data.frame(ev)
  d$WT <- 70
  d$FED <- 0
  d$STUDY_RADIO <- 0
  d$OCC <- occ
  d$SEXF <- sex_draw[d$id]
  d$CONMED_RIFAMPICIN <- 0
  d$CONMED_RIFABUTIN <- rifabutin
  d$treatment <- if (rifabutin == 1) "DTG + rifabutin" else "DTG alone"
  d
}

events_alone <- ddi_arm(0, 1)
events_rfb <- ddi_arm(1, 2)
```

```{r simulate}
solve_arm <- function(d) {
  rxode2::rxSolve(
    ui, d, keep = "treatment", returnType = "data.frame",
    maxsteps = 200000L, atol = 1e-10, rtol = 1e-8
  ) |>
    dplyr::filter(time >= t0) |>
    dplyr::mutate(time = time - t0)
}

# rxSetSeed() is re-issued immediately before each solve so that the two arms
# use common random numbers: the same subject keeps the same clearance eta in
# both arms, which is what makes the per-subject ratio a paired quantity.
rxode2::rxSetSeed(4242)
sim_alone <- solve_arm(events_alone)
rxode2::rxSetSeed(4242)
sim_rfb <- solve_arm(events_rfb)

sim_ddi <- dplyr::bind_rows(sim_alone, sim_rfb) |>
  dplyr::mutate(treatment = factor(treatment,
                                   levels = c("DTG alone", "DTG + rifabutin")))
stopifnot(nrow(sim_ddi) > 0, all(is.finite(sim_ddi$Cc)), all(sim_ddi$Cc >= 0))
```

# Replicate published figures

```{r figure-2, fig.cap = "Replicates Figure 2 of Kawuma 2023."}
sim_ddi |>
  dplyr::group_by(treatment, time) |>
  dplyr::summarise(
    Q10 = stats::quantile(Cc, 0.10),
    Q50 = stats::median(Cc),
    Q90 = stats::quantile(Cc, 0.90),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.18, colour = NA) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 0.064, linetype = "dashed") +
  geom_hline(yintercept = 0.3, linetype = "dotted") +
  scale_y_log10() +
  scale_x_continuous(breaks = seq(0, 24, 4)) +
  labs(
    x = "Time after dose at steady state (h)",
    y = "Dolutegravir concentration (mg/L)",
    colour = NULL, fill = NULL,
    title = "Figure 2 - median simulated steady-state profile, 50 mg once daily",
    caption = paste(
      "Median with 10th-90th percentile band, 70 kg fasted individuals.",
      "Dashed line: PA-IC90 0.064 mg/L. Dotted line: EC90 0.3 mg/L.",
      "Replicates Figure 2 of Kawuma 2023."
    )
  ) +
  theme_bw() +
  theme(legend.position = "top")
```

The rifabutin profile crosses the dolutegravir-alone profile: the smaller
central volume raises the peak and lowers the trough while preserving the area
between them. That crossing is the paper's Figure 2.

# PKNCA validation

```{r pknca}
sim_nca <- sim_ddi |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time = 0 record per (id, treatment). The steady-state grid starts
# at the dosing time, so a row already exists; the bind_rows/distinct pattern is
# kept as a defensive no-op that would supply one if the grid ever changed.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(treatment, id, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- sim_nca |>
  dplyr::distinct(id, treatment) |>
  dplyr::mutate(time = 0, amt = 50)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0, end = 24,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, ctrough = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(treatment, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
stopifnot(nrow(nca_wide) == 2 * n_subj, !anyNA(nca_wide$cmax))
```

## Cohort reproduction of the reported exposure changes

```{r ddi-ratios}
paired <- nca_wide |>
  dplyr::select(id, treatment, cmax, ctrough, auclast) |>
  tidyr::pivot_wider(names_from = treatment,
                     values_from = c(cmax, ctrough, auclast),
                     names_sep = "|")

ratio_summary <- tibble::tibble(
  `NCA parameter` = c("Cmax", "Cmin (Ctrough)", "AUC0-24"),
  `Median change (%)` = c(
    100 * (stats::median(paired$`cmax|DTG + rifabutin` /
                           paired$`cmax|DTG alone`) - 1),
    100 * (stats::median(paired$`ctrough|DTG + rifabutin` /
                           paired$`ctrough|DTG alone`) - 1),
    100 * (stats::median(paired$`auclast|DTG + rifabutin` /
                           paired$`auclast|DTG alone`) - 1)
  ),
  `Kawuma 2023 (%)` = c(15.1, -30.1, 0),
  `Kawuma 2023 95% CI` = c("-49.8 to 268", "-24.7 to -51.5", "no effect")
)

knitr::kable(ratio_summary, digits = 1,
             caption = "Median of the per-subject exposure ratio (rifabutin arm / alone arm) over 200 paired subjects, against the changes reported in Kawuma 2023 Results.")
```

```{r ddi-ratio-gates}
cmax_med <- ratio_summary$`Median change (%)`[1]
cmin_med <- ratio_summary$`Median change (%)`[2]
auc_med <- ratio_summary$`Median change (%)`[3]

stopifnot(
  # These are cohort medians and rxode2 partitions its RNG streams per solver
  # thread, so they move with the machine. Realised across 1 / 2 / 4 / 8 / 16
  # threads at n = 200: Cmax +14.6 to +30.3, Cmin -19.0 to -28.1, AUC -4.8 to
  # +8.8. The bounds sit outside those ranges. They still go red on the
  # failure that matters: siting the rifabutin effect on clearance rather than
  # volume drives Cmax and Cmin both to about -27% and AUC to about -27%.
  cmax_med > 5 && cmax_med < 45,
  cmin_med < -10 && cmin_med > -40,
  abs(auc_med) < 15
)
```

The Cmax spread is wide because between-occasion variability is carried on all
three absorption parameters, so an individual's peak moves far more than the
volume effect alone would suggest -- which is exactly why the paper's own 95%
interval on the Cmax change spans -49.8% to +268%. The 95% range of the
simulated per-subject Cmax ratio is reported below for comparison with that
interval.

```{r cmax-spread}
cmax_ratio <- paired$`cmax|DTG + rifabutin` / paired$`cmax|DTG alone`
sprintf("Simulated 95%% range of the per-subject Cmax ratio: %+.0f%% to %+.0f%%",
        100 * (stats::quantile(cmax_ratio, 0.025) - 1),
        100 * (stats::quantile(cmax_ratio, 0.975) - 1))
```

## Pharmacokinetic target attainment

Kawuma 2023 Results and Discussion:

> in both dosing scenarios, there is a greater than 99% probability that
> dolutegravir trough concentrations are higher than the IC90 of 0.064 mg/L in
> a 70-kg individual and the probability that they are greater than the target
> of 0.3 mg/L is more than 85.9%.

```{r target-attainment}
attainment <- nca_wide |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    `Geometric mean C24 (mg/L)` = exp(mean(log(ctrough))),
    `P(C24 > 0.064 mg/L)` = mean(ctrough > 0.064),
    `P(C24 > 0.3 mg/L)` = mean(ctrough > 0.3),
    .groups = "drop"
  ) |>
  dplyr::rename(Regimen = treatment)

knitr::kable(attainment, digits = 3,
             caption = "Steady-state trough target attainment, dolutegravir 50 mg once daily in fasted 70 kg individuals (n = 200 per arm).")

stopifnot(
  # Absolute bounds taken from the paper's own claims, not from one run.
  # Realised across 1 / 2 / 4 / 8 / 16 threads: P(> 0.064) 0.995-1.000 in both
  # arms; P(> 0.3) 0.895-0.985. Both bounds go red on a mis-sited interaction:
  # applying rifampicin-strength induction instead drops P(> 0.064) to ~0.7
  # and P(> 0.3) to ~0.
  all(attainment$`P(C24 > 0.064 mg/L)` >= 0.98),
  all(attainment$`P(C24 > 0.3 mg/L)` >= 0.80)
)
```

Both claims hold. The simulated probability of clearing the conservative
0.3 mg/L target is a few points above the paper's stated floor of 85.9%, which
is the direction the paper's phrasing ("more than 85.9%") allows; the residual
gap is within the Monte-Carlo spread of a 200-subject cohort (a standard error
of about 2 percentage points at p = 0.9) plus the unstated sex composition of
the authors' own simulated cohort.

# Cross-check against the predecessor model's published simulations

The rifabutin arm is new in this paper, but the rifampicin arm and the
underlying disposition are inherited from Kawuma 2022, whose Table 3 publishes
geometric-mean steady-state trough concentrations for a 70 kg individual under
three regimens. Those numbers come from the **predecessor** model, whose
parameter estimates differ slightly from the ones packaged here (CL/F 1.03 vs
1.03, Vc/F 12.7 vs 13.3, Q/F 0.883 vs 0.675, Vp/F 3.85 vs 3.52 L), so this is a
consistency check across two related models rather than a self-comparison. It
is nonetheless the only published absolute-exposure anchor available for this
model, and it independently exercises the rifampicin effect, the twice-daily
regimen and the dose proportionality.

```{r regimen-crosscheck}
n_cross <- 100
regimens <- tibble::tribble(
  ~treatment,                  ~dose, ~ii, ~rif, ~id_offset,
  "DTG 50 mg OD",                 50,  24,    0,         0L,
  "DTG 50 mg BD + rifampicin",    50,  12,    1,       100L,
  "DTG 100 mg OD + rifampicin",  100,  24,    1,       200L
)

set.seed(20260906)
cross_events <- do.call(dplyr::bind_rows, lapply(
  seq_len(nrow(regimens)),
  function(i) {
    r <- regimens[i, ]
    ev <- rxode2::et(amt = r$dose, ii = r$ii, until = t0 + r$ii, cmt = "depot") |>
      rxode2::et(t0 + obs_grid[obs_grid <= r$ii]) |>
      rxode2::et(id = r$id_offset + seq_len(n_cross))
    d <- as.data.frame(ev)
    d$WT <- 70
    d$FED <- 0
    d$STUDY_RADIO <- 0
    d$OCC <- 1
    d$SEXF <- stats::rbinom(nrow(d), 1, 0.32)
    d$CONMED_RIFAMPICIN <- r$rif
    d$CONMED_RIFABUTIN <- 0
    d$treatment <- r$treatment
    d$tau <- r$ii
    d
  }
))
stopifnot(!anyDuplicated(unique(cross_events[, c("id", "time", "evid")])))

rxode2::rxSetSeed(90210)
sim_cross <- rxode2::rxSolve(
  ui, cross_events, keep = c("treatment", "tau"), returnType = "data.frame",
  maxsteps = 200000L, atol = 1e-10, rtol = 1e-8
) |>
  dplyr::filter(time >= t0) |>
  dplyr::mutate(time = time - t0)

cross_nca <- sim_cross |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)
cross_conc <- PKNCA::PKNCAconc(cross_nca, Cc ~ time | treatment + id)
cross_dose_df <- do.call(dplyr::bind_rows, lapply(
  seq_len(nrow(regimens)),
  function(i) {
    r <- regimens[i, ]
    tibble::tibble(id = r$id_offset + seq_len(n_cross),
                   treatment = r$treatment, time = 0, amt = r$dose)
  }
))
stopifnot(nrow(cross_dose_df) == 3 * n_cross)
cross_dose <- PKNCA::PKNCAdose(cross_dose_df, amt ~ time | treatment + id)
cross_intervals <- sim_cross |>
  dplyr::distinct(treatment, tau) |>
  dplyr::mutate(start = 0, end = tau, ctrough = TRUE, cmax = TRUE) |>
  dplyr::select(start, end, ctrough, cmax, treatment)

cross_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(cross_conc, cross_dose,
                                            intervals = cross_intervals))

published_2022 <- tibble::tribble(
  ~treatment,                   ~ctrough,
  "DTG 50 mg OD",                  0.878,
  "DTG 50 mg BD + rifampicin",     0.608,
  "DTG 100 mg OD + rifampicin",    0.220
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = cross_res,
  reference = published_2022,
  by = "treatment",
  params = "ctrough",
  units = c(ctrough = "mg/L"),
  tolerance_pct = 20
)

knitr::kable(cmp, digits = 3,
             caption = "Simulated steady-state trough (median over 100 subjects per regimen, 70 kg, fasted) against the geometric-mean troughs published in Kawuma 2022 Table 3 for the predecessor model. * marks a difference greater than 20%.")
```

```{r crosscheck-gate}
cross_wide <- as.data.frame(cross_res) |>
  dplyr::filter(PPTESTCD == "ctrough") |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(sim = stats::median(PPORRES), .groups = "drop") |>
  dplyr::left_join(published_2022, by = "treatment") |>
  dplyr::mutate(pct = 100 * (sim / ctrough - 1))

stopifnot(
  nrow(cross_wide) == 3, !anyNA(cross_wide$pct),
  # A 25% band. This compares a median across a 100-subject cohort of THIS
  # model against a geometric mean published for the PREDECESSOR model, so it
  # can never be tight; it is here to catch a broken rifampicin effect (which
  # would put the two rifampicin arms out by a factor of ~2.4) or a lost dose
  # proportionality (a factor of 2 on the 100 mg arm).
  all(abs(cross_wide$pct) < 25)
)
```

All three regimens land within 25% of the predecessor model's published
geometric means, with the two rifampicin arms confirming that the +143%
clearance effect and the twice-daily accumulation are both encoded correctly.

# Assumptions and deviations

* **Allometric exponents are not printed in the 2023 paper.** Table 1
  footnote b states only that allometric scaling with weight (70 kg median) was
  applied to CL, Q, Vc and Vp. The exponent values -- 0.75 for the clearances
  and 1 for the volumes, both fixed rather than estimated -- come from the
  Methods of the predecessor paper the model extends (Kawuma 2022,
  doi:10.1128/aac.00215-22, "Pharmacokinetic analysis"). They are the only
  numeric values in the model file that are not from the 2023 paper's own
  Table 1, and both are wrapped in `fixed()`.
* **The `%CV` convention is the paper's own.** Table 1 footnote e states
  `CV% = sqrt(omega^2) * 100`, i.e. the tabulated percentage is 100 times the
  log-scale standard deviation. The variances in `ini()` are therefore
  `(CV% / 100)^2` and **not** `log(1 + CV^2)`. The two conventions differ by
  under 4% at the largest reported variability here (60.7%), but the footnote
  is explicit, so the direct reading is used.
* **Between-occasion variability is encoded as four occasions.** Kawuma 2022
  defines an occasion as a dosing event leading to at least one observation,
  with each sampling visit contributing a predose and a postdose occasion, and
  does not fix an occasion count. A single variance per absorption parameter is
  reported and shared across occasions, so occasion 1 carries the estimated
  variance and occasions 2-4 fix it to the same value (the registered
  `$OMEGA BLOCK(1) SAME` idiom). Extending the chain past four occasions is a
  mechanical copy of the `fix()` lines. Pass `OCC = 1` for single-occasion
  records.
* **Study and prandial state are perfectly confounded in the source data, and
  are split across two covariates here.** RADIO dosed fed and NCT01231542
  dosed fasted, so a single study indicator could carry both the absorption-lag
  difference and the residual-error difference. The lag time is carried by the
  general `FED` indicator, because the authors label the two lag-time rows by
  prandial state (Table 1 footnotes c and d), index the lag by prandial state
  when simulating (Kawuma 2022 Figure 3 caption) and attribute the difference
  mechanistically to food (Kawuma 2022 Discussion). The additive residual error
  is carried by `STUDY_RADIO`, because it tracks the two studies' assays: the
  additive term was constrained to at least 20% of each study's LLOQ (0.020
  mg/L for NCT01231542, 0.050 mg/L for RADIO), and the printed lower confidence
  bounds of 0.00435 and 0.0103 mg/L sit on exactly those floors. A user who
  wants the paper's literal study effect should set both indicators together.
* **The sex effect keeps the paper's female reference.** Kawuma 2023 reports
  "Male sex on `ka` = -38.1%", so the printed typical `ka` of 1.63 /h is the
  value for women. The canonical `SEXF` column is 1 = female, so the
  coefficient is applied to `(1 - SEXF)`, preserving both the register's
  orientation and the paper's verbatim numbers.
* **Rifabutin's effect is encoded on `Vc` only.** The authors fitted rifabutin
  on clearance (+36%), on bioavailability alone (not significant) and jointly on
  clearance and bioavailability (+41% and +32%), and selected the volume effect
  on goodness-of-fit, plausibility and BIC (-870.7 vs -866.1). Only the final,
  selected model is packaged, per the library's replicate-the-authors'-structure
  policy; the rejected alternatives are recorded here rather than as extra
  parameters.
* **Simulated cohort composition.** The paper states only that 1000 individuals
  of 70 kg were simulated. Sex is drawn here at the studies' 68% male / 32%
  female split, doses are fasted (`FED = 0`) following the predecessor paper's
  simulation convention, and cohorts are capped at 200 per arm by library
  policy rather than the paper's 1000. Sex enters the model only through `ka`
  and has a negligible effect on a steady-state trough.
* **The INSPIRING patient cohort is not represented.** The predecessor paper
  externally validated against INSPIRING but could not reliably co-model
  healthy-volunteer and patient data; the packaged model is the
  healthy-volunteer model, which the authors note slightly underpredicts
  troughs in patients, making its target-attainment predictions conservative.
* **The cross-check in the previous section uses a different model's published
  values.** Kawuma 2023 publishes no absolute NCA table, so the only available
  absolute-exposure anchor is Kawuma 2022 Table 3, generated with the
  predecessor parameter set. That comparison is gated at 25% and is documented
  as a cross-model consistency check, not a self-validation.
* **No erratum applies.** Europe PMC records exactly one linked item for
  doi:10.1111/bcp.15604, typed "Comment in": a letter by Theile, Nilles and Meid
  (Br J Clin Pharmacol 2023;89(7):2329-2331, doi:10.1111/bcp.15740) arguing from
  physiologically based modelling and in vitro data that rifabutin alters
  dolutegravir kinetics through simultaneous P-glycoprotein induction *and*
  inhibition. That is a mechanistic commentary supporting the authors' own
  transporter hypothesis, not a correction: it revises no parameter estimate and
  the packaged values are those of the original publication. No erratum,
  corrigendum or author correction is indexed for this article.
