---
title: "Tenofovir from TDF or TAF (Kawuma 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tenofovir from TDF or TAF (Kawuma 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5, dpi = 100
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Kawuma_2023_tenofovir"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Kawuma_2023_tenofovir"))$description`
* Article: <https://doi.org/10.1002/psp4.12955>

Tenofovir disoproxil fumarate (TDF) and tenofovir alafenamide (TAF) are both
prodrugs of tenofovir. TDF converts to tenofovir largely in plasma, whereas TAF
is taken up into cells and converted there, so TAF can be given at a much lower
dose and produces roughly ten-fold lower plasma tenofovir concentrations.

Kawuma 2023 is notable for fitting **one** tenofovir disposition model to both
prodrugs. The authors argue explicitly against the earlier approach of Greene et
al., who fitted TDF and TAF separately and reported two clearance values
differing about ten-fold: once tenofovir has reached plasma it should be
distributed and eliminated the same way regardless of which prodrug delivered
it. In this model the ten-fold exposure difference is produced entirely by the
input function, not by the disposition.

Consequently the model has three input compartments and one shared two-
compartment disposition:

| Compartment | Fed by | Rate out |
|---|---|---|
| `depot`  | a TDF dose | `Ka_TDF` = 3.04 /h |
| `depot2` | the fast fraction of a TAF dose | `Ka_TAF` = 1.45 /h |
| `depot3` | the slow fraction of a TAF dose, sequestered intracellularly | half-life 6.83 days |

# Doses are tenofovir-equivalent amounts

The model's dosing unit is **mg of tenofovir**, not mg of prodrug. Kawuma 2023
Methods converts by molecular weight (tenofovir 287.2 g/mol; TDF 635.5 g/mol;
TAF 476.5 g/mol):

```{r doses}
mw_tfv <- 287.2
mw_tdf <- 635.5
mw_taf <- 476.5

# The paper's own rounded conversions (Methods): 300 mg TDF -> 136 mg
# tenofovir; 25 mg TAF -> 15 mg tenofovir.
tfv_tdf300 <- 136
tfv_taf25 <- 15
# 10 mg TAF is simulated in Table 3 but its conversion is not printed.
tfv_taf10 <- 10 * mw_tfv / mw_taf

c(
  check_tdf300 = 300 * mw_tfv / mw_tdf,
  check_taf25 = 25 * mw_tfv / mw_taf,
  taf10 = tfv_taf10
)
```

`r sprintf("%.1f", 300 * mw_tfv / mw_tdf)` and
`r sprintf("%.2f", 25 * mw_tfv / mw_taf)` round to the paper's 136 mg and 15 mg,
confirming the conversion.

# Population

The model was built on 279 tenofovir concentrations from 41 South African
adults living with HIV, enrolled in a pharmacokinetic substudy nested within the
ADVANCE trial (NCT03122262). Twenty-one participants received TDF 300 mg and
twenty received TAF 25 mg, each co-formulated with emtricitabine 200 mg and
given with dolutegravir 50 mg once daily. Sampling was rich -- predose and at 1,
2, 4, 6, 8 and 24 h postdose -- and was performed after at least 48 weeks of
treatment, so every profile is at steady state.

Baseline characteristics (Kawuma 2023 Table 1, "ALL" column): median (IQR) age
31.0 (29.0-36.0) years, weight 73.1 (67.2-85.2) kg, height 167 (161-174) cm,
and Cockcroft-Gault creatinine clearance at screening 120 (96.0-140) mL/min.
27 of 41 participants (65.9%) were female. None of the 279 samples fell below
the 0.0005 mg/L limit of quantification.

The same information is available programmatically:

```{r population}
str(rxode2::rxode2(readModelDb("Kawuma_2023_tenofovir"))$population)
```

# Source trace

Every value in `ini()` and every non-obvious equation in `model()`, with the
place in Kawuma 2023 it came from.

| Model element | Value | Source |
|---|---|---|
| `lcl` | CL 44.7 L/h (40.2-49.5) | Table 2, `CL (L/h)` row |
| `lvc` | Vc 378 L (319-459) | Table 2, `Vc (L)` row |
| `lq` | Q 157 L/h (103-233) | Table 2, `Q (L/h)` row |
| `lvp` | Vp 356 L (298-438) | Table 2, `Vp (L)` row |
| `e_wt_cl_q` | 0.75, fixed | Methods: "allometric exponents for clearance and volume were fixed to 0.75 and 1" |
| `e_wt_vc_vp` | 1, fixed | Methods, as above |
| reference weight | 70 kg | Table 2 footnote c; Results "For a typical individual of 70 kg" |
| `lka_tdf` | Ka_TDF 3.04 /h (2.11-3.88) | Table 2, `Ka_TDF (1/h)` row |
| `lfdepot_tdf` | F_TDF 1, fixed | Table 2, `F_TDF` row, "1-Fixed" |
| `lfdepot_taf` | F_TAF 0.822 (0.723-0.939) | Table 2, `F_TAF` row; Results text repeats 82.2% (72.3-93.9) |
| `logitffo` | Frac_TAF-Fast 32.4% (27.0-37.7) | Table 2, `Frac_TAF-Fast (%)` row |
| `lka_taf_fast` | Ka_TAF 1.45 /h (0.924-2.60) | Table 2, `Ka_TAF (1/h)` row |
| `lka_taf_slow` | t1/2 6.83 days, fixed; rate = log(2)/(6.83 x 24) /h | Table 2, `t1/2_TAF-Slow, days` row, "6.83-Fixed" |
| `etalcl` | BSV on CL 20.1% (16.1-24.7) | Table 2 CL row, variability column, footnote d |
| `etaiov_fdepot_tdf_*` | BOV on F_TDF 23.9% (18.2-30.3) | Table 2 F_TDF row, footnote e |
| `etaiov_ka_tdf_*` | BOV on Ka_TDF 114.5% (68.4-162) | Table 2 Ka_TDF row, footnote e |
| `etaiov_ka_taf_*` | BOV on Ka_TAF 66.3% (31.0-91.6) | Table 2 Ka_TAF row, footnote e |
| variance scale | omega^2 = (CV%/100)^2 | Table 2 footnote b: "calculated by CV% = sqrt(omega^2) x 100" |
| `propSd` | 11.9% (10.8-13.4) | Table 2, `Proportional error (%)` row |
| `addSd` | 0.2 x 0.0005 = 0.0001 mg/L, fixed | Table 2 `Additive error` row "20% of LLOQ-Fixed"; footnote f "The LLOQ was 0.0005 mg/L" |
| two parallel TAF pathways | structure | Figure 2 legend; Results paragraph 3 |
| two-compartment disposition | structure | Results: "A two-compartment disposition model (dOFV = -47...) best described" |
| occasion definition | 2 occasions | Methods: "an 'occasion' was defined as a dose with its proceeding sample" |
| dose conversion | 136 mg / 15 mg tenofovir | Methods, "Population pharmacokinetic modeling" paragraph 1 |

Note the variance scale. Table 2 footnote b defines `CV% = sqrt(omega^2) * 100`,
i.e. the printed percentages are omega on the standard-deviation scale directly.
The variances in `ini()` are therefore `(CV/100)^2` and **not** the exp-based
`log(1 + CV^2)` conversion that many popPK papers imply.

# How a dose is given to this model

A **TDF** regimen doses `depot` only. A **TAF** regimen doses **both** `depot2`
and `depot3` with the same tenofovir-equivalent amount; the `f()` terms in the
model split that amount between the fast and slow pathways so that the two
records together deliver `F_TAF` worth of one dose:

```
f(depot2) = F_TAF * Frac_TAF-Fast
f(depot3) = F_TAF * (1 - Frac_TAF-Fast)
```

```{r events}
mod <- rxode2::rxode2(readModelDb("Kawuma_2023_tenofovir"))

#' Build an event table for one arm.
#'
#' @param tfv_dose tenofovir-equivalent dose in mg
#' @param arm "TDF" or "TAF"
#' @param ndays number of once-daily doses
#' @param obs_from,obs_to,obs_by observation grid, in hours
build_events <- function(tfv_dose, arm, ndays, obs_from, obs_to, obs_by) {
  last_dose <- 24 * (ndays - 1)
  ev <-
    if (arm == "TDF") {
      rxode2::et(amt = tfv_dose, cmt = "depot", ii = 24, until = last_dose)
    } else {
      rxode2::et(amt = tfv_dose, cmt = "depot2", ii = 24, until = last_dose) |>
        rxode2::et(amt = tfv_dose, cmt = "depot3", ii = 24, until = last_dose)
    }
  # Observe on the ODE state `central`; rxode2 returns the algebraic
  # observable Cc as a column at those records.
  rxode2::et(ev, seq(obs_from, obs_to, by = obs_by), cmt = "central")
}
```

Both arms are solved with `OCC = 0`, which zeroes every occasion indicator and
so removes the between-occasion contribution. This matches how the authors ran
their own simulations (Table 3 footnote a: "excluding between-occasion
variability on absorption parameters").

# Structure check: where the tenofovir comes from

A single 25 mg TAF dose, split into the two pathways, shows the mechanism the
paper describes: a fast peak within the first few hours from `depot2`, then a
long, almost flat tail sustained by slow release from the intracellular
reservoir `depot3`.

```{r single-dose-taf, fig.cap = "Single-dose tenofovir after TAF 25 mg, typical 70 kg individual, showing the amount remaining in each TAF input compartment. Replicates the mechanism drawn in Figure 2 of Kawuma 2023."}
ev_single <- build_events(tfv_taf25, "TAF", ndays = 1,
                          obs_from = 0, obs_to = 24 * 21, obs_by = 0.5)
d_single <- as.data.frame(ev_single)
d_single$WT <- 70
d_single$OCC <- 0

sim_single <-
  rxode2::rxSolve(rxode2::zeroRe(mod), d_single, returnType = "data.frame")

sim_single |>
  filter(!is.na(Cc)) |>
  select(time, depot2, depot3) |>
  pivot_longer(c(depot2, depot3), names_to = "compartment",
               values_to = "amount_mg") |>
  mutate(compartment = recode(compartment,
                              depot2 = "depot2 (fast pathway)",
                              depot3 = "depot3 (intracellular reservoir)")) |>
  ggplot(aes(time / 24, amount_mg, colour = compartment)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Days since dose", y = "Tenofovir remaining (mg)",
       colour = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

The fast pathway is essentially exhausted within a day while the reservoir is
still nearly full, which is what produces tenofovir's much longer apparent
plasma half-life after TAF than after TDF. Ruane et al. (cited in the Kawuma
2023 Discussion) measured 14.86 h after TDF versus 40.19 h after TAF; here that
difference is generated by the input function while disposition is held
identical.

```{r pathway-split}
# Fraction of the administered tenofovir amount routed down each pathway.
# The model applies f() at the dose record, so the amount in each depot at
# t = 0+ is the check.
t0 <- sim_single |> filter(!is.na(Cc)) |> slice_min(time, n = 1)
c(
  fast_mg = t0$depot2,
  slow_mg = t0$depot3,
  total_mg = t0$depot2 + t0$depot3,
  expected_total_mg = tfv_taf25 * 0.822,
  fast_fraction = t0$depot2 / (t0$depot2 + t0$depot3)
)

stopifnot(
  # The two records together deliver exactly F_TAF of the dose ...
  abs((t0$depot2 + t0$depot3) - tfv_taf25 * 0.822) < 1e-6,
  # ... split in the published 32.4 / 67.6 ratio.
  abs(t0$depot2 / (t0$depot2 + t0$depot3) - 0.324) < 1e-6
)
```

# Steady state takes weeks on TAF, days on TDF

The reservoir's 6.83-day half-life means a TAF regimen needs roughly five of
those half-lives -- about 34 days -- before the slow pathway stops accumulating.
TDF, with no reservoir, reaches steady state within a few days. We therefore run
45 once-daily doses and take the final dosing interval for both arms, so the
same run-in covers the slower arm.

```{r approach-to-ss, fig.cap = "Approach to steady state on once-daily dosing, typical 70 kg individual. Tenofovir from TAF accumulates over about a month because of the intracellular reservoir; tenofovir from TDF plateaus within days."}
n_days_ss <- 45

approach <- bind_rows(
  lapply(
    list(
      list(arm = "TDF", dose = tfv_tdf300),
      list(arm = "TAF", dose = tfv_taf25)
    ),
    function(a) {
      d <- as.data.frame(
        build_events(a$dose, a$arm, ndays = n_days_ss,
                     obs_from = 0, obs_to = 24 * n_days_ss, obs_by = 3)
      )
      d$WT <- 70
      d$OCC <- 0
      out <- rxode2::rxSolve(rxode2::zeroRe(mod), d, returnType = "data.frame")
      out$arm <- a$arm
      out[!is.na(out$Cc), c("time", "Cc", "arm")]
    }
  )
)

ggplot(approach, aes(time / 24, Cc * 1000)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~arm, scales = "free_y") +
  labs(x = "Days", y = "Plasma tenofovir (ng/mL)") +
  theme_bw()
```

```{r ss-attained}
# Trough on the last day vs the day before: both arms must have plateaued.
trough_ratio <- approach |>
  filter(time %in% c(24 * (n_days_ss - 1), 24 * (n_days_ss - 2))) |>
  group_by(arm) |>
  summarise(ratio = Cc[time == max(time)] / Cc[time == min(time)],
            .groups = "drop")
trough_ratio

stopifnot(all(abs(trough_ratio$ratio - 1) < 0.01))
```

# Replicating the published simulated AUC (Table 3)

Kawuma 2023 Table 3 compares AUC0-24h simulated with the final model against
values reported by three other studies. The simulated column is the model
prediction we can check directly. The authors state their AUCs were obtained as
`AUC0-24h = F_i * Dose_i / CL_i` for "1000 runs of a single typical individual
of the weight and dose specified per study excluding between-occasion
variability on absorption parameters", so the target is a typical-value
prediction. We therefore simulate with `zeroRe()` and compute the AUC over the
final dosing interval by NCA -- an independent route to the same number, which
also exercises the ODE mass balance, the `f()` splits and the allometry.

```{r ss-scenarios}
scenarios <- tibble::tribble(
  ~scenario,                ~arm,  ~wt,  ~tfv_dose,   ~auclast_ref,
  "TDF 300 mg, 77 kg",      "TDF", 77.0, tfv_tdf300,  2795,
  "TDF 300 mg, 52 kg",      "TDF", 52.0, tfv_tdf300,  3761,
  "TAF 25 mg, 77 kg",       "TAF", 77.0, tfv_taf25,   253.6,
  "TAF 25 mg, 78.7 kg",     "TAF", 78.7, tfv_taf25,   259.8,
  "TAF 10 mg, 77 kg",       "TAF", 77.0, tfv_taf10,   101.6,
  "TAF 10 mg, 74.3 kg",     "TAF", 74.3, tfv_taf10,   108.6
)
knitr::kable(
  scenarios |> select(-arm) |> rename("Scenario" = scenario,
                                      "Weight (kg)" = wt,
                                      "Tenofovir dose (mg)" = tfv_dose,
                                      "Published AUC0-24 (ng*h/mL)" = auclast_ref),
  digits = 2,
  caption = "Simulation scenarios and the corresponding published model-simulated AUC0-24h, from Kawuma 2023 Table 3."
)
```

```{r ss-simulate}
start_ss <- 24 * (n_days_ss - 1)
end_ss <- 24 * n_days_ss

# One rxSolve call per scenario; solving an rxUi scales poorly when many
# subjects go through a single call.
ss_conc <- bind_rows(lapply(seq_len(nrow(scenarios)), function(i) {
  s <- scenarios[i, ]
  d <- as.data.frame(
    build_events(s$tfv_dose, s$arm, ndays = n_days_ss,
                 obs_from = start_ss, obs_to = end_ss, obs_by = 0.1)
  )
  d$WT <- s$wt
  d$OCC <- 0
  out <- rxode2::rxSolve(rxode2::zeroRe(mod), d, returnType = "data.frame")
  out <- out[!is.na(out$Cc), ]
  data.frame(
    scenario = s$scenario,
    id = 1L,
    time = out$time,
    # mg/L -> ng/mL so the NCA output is directly in the paper's units
    conc = out$Cc * 1000
  )
}))

ss_dose <- scenarios |>
  transmute(scenario, id = 1L, time = start_ss, dose = tfv_dose)
```

```{r ss-nca}
conc_obj <- PKNCA::PKNCAconc(
  ss_conc, conc ~ time | scenario + id, concu = "ng/mL", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  ss_dose, dose ~ time | scenario + id, doseu = "mg"
)

intervals <- data.frame(
  start = start_ss, end = end_ss,
  auclast = TRUE, cmax = TRUE, tmax = TRUE, cmin = TRUE, cav = TRUE
)

ss_nca <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)
```

```{r ss-comparison}
ref_tbl <- scenarios |> transmute(scenario, auclast = auclast_ref)

cmp <- ncaComparisonTable(
  simulated = ss_nca,
  reference = ref_tbl,
  by = "scenario",
  params = "auclast",
  units = c(auclast = "ng*h/mL")
)
knitr::kable(
  cmp, digits = 1,
  caption = "Simulated steady-state AUC0-24h versus the model-simulated values published in Kawuma 2023 Table 3."
)
attr(cmp, "footnote")
```

```{r ss-comparison-assert}
# ncaComparisonTable() returns display-formatted character columns, so the
# assertion is computed from the underlying NCA values.
auc_sim <- as.data.frame(ss_nca$result) |>
  filter(PPTESTCD == "auclast") |>
  select(scenario, auc_nca = PPORRES)

pct_diff <- scenarios |>
  left_join(auc_sim, by = "scenario") |>
  mutate(pct = 100 * (auc_nca - auclast_ref) / auclast_ref) |>
  pull(pct)
summary(pct_diff)

stopifnot(
  # All six scenarios are typical-value predictions with no random component
  # on our side, so the only spread is the Monte-Carlo noise in the paper's
  # own 1000-draw simulation. 8% is comfortably above that and well below
  # anything a mis-transcribed parameter, dose or unit would produce.
  max(abs(pct_diff)) < 8
)
```

Every scenario reproduces to within a few percent, across a four-fold range of
tenofovir dose (6 to 136 mg), both prodrugs, and weights from 52 to 78.7 kg.

# The AUC identity that the disposition model implies

Because elimination is linear and every dose eventually reaches plasma, the
steady-state AUC over a dosing interval must equal `F * Dose / CL`. This is a
useful independent check: it goes red if the `f()` splits do not sum to `F_TAF`,
if the allometric term is misapplied, or if the run-in is too short for the
slow pathway to have reached steady state.

```{r closed-form}
closed_form <- scenarios |>
  mutate(
    cl = 44.7 * (wt / 70)^0.75,
    fbio = if_else(arm == "TDF", 1, 0.822),
    auc_closed = 1000 * fbio * tfv_dose / cl
  )

auc_sim <- as.data.frame(ss_nca$result) |>
  filter(PPTESTCD == "auclast") |>
  select(scenario, auc_nca = PPORRES)

closed_chk <- closed_form |>
  left_join(auc_sim, by = "scenario") |>
  mutate(pct_diff = 100 * (auc_nca - auc_closed) / auc_closed)

knitr::kable(
  closed_chk |>
    select(scenario, auc_closed, auc_nca, pct_diff) |>
    rename("Scenario" = scenario,
           "F x Dose / CL (ng*h/mL)" = auc_closed,
           "NCA of simulated profile (ng*h/mL)" = auc_nca,
           "% difference" = pct_diff),
  digits = 2,
  caption = "Steady-state AUC0-24h from NCA of the simulated profile against the closed-form F x Dose / CL identity."
)

stopifnot(
  # Both sides use the same parameters, so this is deterministic: no cohort
  # is drawn and the bound does not depend on an RNG stream. The TDF rows
  # agree to 0.03%. The TAF rows sit about 0.8% low because 45 days is 6.6
  # half-lives of the 6.83-day reservoir, so roughly 2^-6.6 = 1% of the slow
  # pathway's steady-state accumulation is still missing -- a real, and
  # quantitatively predictable, structural shortfall rather than integration
  # error. The bound is set just above it.
  max(abs(closed_chk$pct_diff)) < 1.5
)
```

# Cohort simulation and steady-state exposure spread

The parameter table supports one source of between-subject variability, on
clearance (20.1% CV). Below, 200 subjects per arm are simulated at the cohort's
weight distribution to show the resulting steady-state exposure spread and the
ten-fold separation between the arms.

```{r cohort}
rxode2::rxSetSeed(20230601)
set.seed(20230601)

n_sub <- 200

# Weight distribution matched to Table 1: median 73.1 kg, IQR 67.2-85.2 kg.
# A log-normal whose median is 73.1 and whose IQR matches the reported one
# gives sdlog = log(85.2 / 67.2) / (2 * qnorm(0.75)).
sdlog_wt <- log(85.2 / 67.2) / (2 * stats::qnorm(0.75))
wt_cohort <- stats::rlnorm(n_sub, meanlog = log(73.1), sdlog = sdlog_wt)

round(stats::quantile(wt_cohort, c(0.25, 0.5, 0.75)), 1)
```

```{r cohort-simulate}
cohort <- bind_rows(lapply(
  list(list(arm = "TDF", dose = tfv_tdf300), list(arm = "TAF", dose = tfv_taf25)),
  function(a) {
    ev <- build_events(a$dose, a$arm, ndays = n_days_ss,
                       obs_from = start_ss, obs_to = end_ss, obs_by = 0.25)
    d <- as.data.frame(ev)
    # Replicate the event table per subject and attach the covariates on the
    # materialised data frame -- assignments onto an rxEt object are dropped.
    d <- do.call(rbind, lapply(seq_len(n_sub), function(j) {
      dj <- d
      dj$id <- j
      dj$WT <- wt_cohort[j]
      dj$OCC <- 0
      dj
    }))
    # rxSetSeed inside the loop so each arm draws its own etas reproducibly.
    rxode2::rxSetSeed(20230601)
    out <- rxode2::rxSolve(mod, d, returnType = "data.frame")
    out <- out[!is.na(out$Cc), ]
    data.frame(arm = a$arm, id = out$id, time = out$time, conc = out$Cc * 1000)
  }
))
```

```{r cohort-plot, fig.cap = "Simulated steady-state tenofovir concentration-time profiles over one dosing interval, 200 subjects per arm. Median and 5th / 95th percentiles, the same summary shown in the visual predictive check of Kawuma 2023 Figure 3. Note the free y scales: TDF exposure is about ten-fold higher."}
cohort |>
  mutate(tad = time - start_ss) |>
  group_by(arm, tad) |>
  summarise(
    med = stats::median(conc),
    lo = stats::quantile(conc, 0.05),
    hi = stats::quantile(conc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, med)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~arm, scales = "free_y") +
  labs(x = "Time after dose (h)", y = "Plasma tenofovir (ng/mL)") +
  theme_bw()
```

```{r cohort-nca}
cohort_conc <- PKNCA::PKNCAconc(
  cohort, conc ~ time | arm + id, concu = "ng/mL", timeu = "h"
)
cohort_dose <- cohort |>
  distinct(arm, id) |>
  mutate(
    dose = if_else(arm == "TDF", tfv_tdf300, tfv_taf25),
    time = start_ss
  )
cohort_dose_obj <- PKNCA::PKNCAdose(
  cohort_dose, dose ~ time | arm + id, doseu = "mg"
)

cohort_nca <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(cohort_conc, cohort_dose_obj, intervals = intervals)
)

cohort_summary <- as.data.frame(cohort_nca$result) |>
  filter(PPTESTCD %in% c("auclast", "cmax", "cmin", "tmax")) |>
  group_by(arm, PPTESTCD) |>
  summarise(
    median = stats::median(PPORRES),
    p05 = stats::quantile(PPORRES, 0.05),
    p95 = stats::quantile(PPORRES, 0.95),
    .groups = "drop"
  )

knitr::kable(
  cohort_summary |>
    rename("Arm" = arm, "Parameter" = PPTESTCD, "Median" = median,
           "5th percentile" = p05, "95th percentile" = p95),
  digits = 2,
  caption = "Simulated steady-state NCA parameters by arm (AUC0-24 and concentrations in ng*h/mL and ng/mL, tmax in h)."
)
```

```{r cohort-ratio}
med_auc <- cohort_summary |>
  filter(PPTESTCD == "auclast") |>
  select(arm, median) |>
  tidyr::pivot_wider(names_from = arm, values_from = median)

tdf_taf_ratio <- med_auc$TDF / med_auc$TAF
tdf_taf_ratio

stopifnot(
  # Kawuma 2023 Figure 1 legend: "when given as TDF, tenofovir in the plasma
  # is more than 10-fold higher than when TAF is administered".
  tdf_taf_ratio > 10
)
```

The observed AUC0-24h in the study itself (Kawuma 2023 Table 3, "Observed (this
study)") was 3022 `ng*h/mL` on TDF and 267.1 `ng*h/mL` on TAF -- a ratio of
`r sprintf("%.1f", 3022 / 267.1)` -- against the
`r sprintf("%.1f", tdf_taf_ratio)`-fold ratio simulated here at the cohort
weight distribution.

# Assumptions and deviations

* **Doses are tenofovir-equivalent amounts, not prodrug amounts.** A user
  dosing this model with `amt = 300` for TDF, or `amt = 25` for TAF, will
  over-predict exposure by roughly the molecular-weight ratio. Use 136 mg and
  15 mg, or convert with the molecular weights given above.

* **A TAF dose requires two dose records**, into `depot2` and `depot3`, both
  carrying the same tenofovir-equivalent amount. The `f()` terms split that
  amount; giving only one record delivers only that pathway's share.

* **Between-occasion variability is encoded with explicit per-occasion etas.**
  rxode2 has no NONMEM-style `| occasion` random-effect level, so the BOV on
  `F_TDF`, `Ka_TDF` and `Ka_TAF` is expanded into one eta per occasion,
  selected by the `OCC` indicators. Kawuma 2023 defines exactly two occasions
  (the unobserved dose the day before the visit, and the observed dose on the
  day of the visit), so two are encoded, with the second variance fixed to the
  first -- the equivalent of NONMEM's `$OMEGA BLOCK(1) SAME`. Because the
  occasion effect enters as an indicator-weighted sum rather than as a bare
  term, rxode2 emits `some etas defaulted to non-mu referenced` on every solve.
  This is expected and affects only mu-referencing speed-ups in a future
  *estimation* run; simulation, which is what this model is shipped for, is
  unaffected. Passing `OCC = 0` (as every simulation here does) zeroes all
  indicators and removes the between-occasion contribution, reproducing the
  authors' own simulation setting.

* **The variance scale is the SD convention.** Table 2 footnote b states
  `CV% = sqrt(omega^2) * 100`, so the `ini()` variances are `(CV/100)^2`. Using
  the more common `log(1 + CV^2)` conversion instead would inflate the BOV on
  `Ka_TDF` (114.5% CV) substantially.

* **`t1/2_TAF-Slow` is fixed, and only weakly identified.** The authors
  estimated it first, found it poorly identifiable, and reported that
  likelihood profiling gave "only small changes in terms of goodness-of-fit"
  anywhere in the range 5 to 60 days. They fixed it at 6.83 days, the published
  intracellular TFV-DP decay half-life. It is encoded here as a fixed
  first-order rate constant `log(2) / (6.83 x 24)` per hour. The sensitivity
  analysis behind this choice is in the paper's Table S1 and Figure S4, which
  are not on disk (see below); the final fixed value itself is printed in
  Table 2 of the main text.

* **The model does not reproduce the secondary TAF peak.** The Discussion notes
  that the observed TAF profiles show "after an initial peak at about 1 h
  postdose, ... a secondary peak at about 4 h postdose", and speculates that it
  reflects intracellular TAF conversion. The fitted structure releases from the
  reservoir by a single slow first-order process with a multi-day half-life,
  which is monotone and cannot produce a secondary peak. This is a property of
  the published model, not of the encoding.

* **Age and creatinine clearance were screened and dropped.** Neither improved
  the fit, and no coefficient is published for either, so neither can be
  encoded. They are recorded in `covariatesDataExcluded` to preserve the
  provenance of the covariate screen. The authors flag the negative renal
  finding as a limitation, attributing it to the narrow creatinine-clearance
  range in the cohort and to creatinine not being measured at the time of drug
  sampling.

* **No variability on Vc, Q, Vp, F_TAF or Frac_TAF-Fast.** Table 2 leaves those
  variability cells blank, and the Results state that within the TAF arm the
  model "did not support the estimation of any variability (BOV or BSV) on the
  bioavailability parameter". No variance is invented for them.

* **Race and ethnicity are not reported.** Table 1 gives sex, age, weight,
  height and creatinine clearance only, so `population$race_ethnicity` records
  the absence rather than a distribution. The weight distribution used for the
  cohort simulation above is a log-normal matched to the published median and
  IQR, not a resampling of the actual 41 participants, whose individual
  covariates are not available in a form on disk.

* **Fat-free mass was tested and rejected.** The authors report that FFM
  scaling did not improve the model over no allometry (dOFV = -3.80) whereas
  total body weight did (dOFV = -18), so total body weight is the size
  descriptor encoded here.

## Supplementary material not obtained

The article's Supporting Information -- Figures S1-S4 (raw data and
goodness-of-fit plots, plus the `t1/2_TAF-Slow` sensitivity figure), Table S1
(the `t1/2_TAF-Slow` sensitivity analysis), a NONMEM control stream deposited as
`PSP4-12-821-s007.mod`, and the concentration dataset
`PSP4-12-821-s006.xlsx` -- could not be retrieved. PMC now serves supplementary
blobs behind a JavaScript proof-of-work challenge and the Wiley landing page
returns HTTP 403.

No parameter used by this model comes from those files: every structural
element is specified in the Methods and the Figure 2 legend, and every final
estimate is printed in Table 2 of the main text. The control stream would be a
useful independent confirmation of the encoding, and is worth retrieving if a
reviewer wants one; note that a control stream's `$THETA` block holds initial,
not final, estimates, so Table 2 would remain the authoritative source either
way.
