---
title: "Vaginal dapivirine ring and film PBPK (Kay 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vaginal dapivirine ring and film PBPK (Kay 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
```

```{r libraries, message = FALSE}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Kay et al. (2018, *Br J Clin Pharmacol* 84(9):1950-1969,
[doi:10.1111/bcp.13625](https://doi.org/10.1111/bcp.13625), PMCID
PMC6089833) built the first physiologically-based pharmacokinetic model
of the **vaginal space**, and applied it to dapivirine (DPV), a
non-nucleoside reverse transcriptase inhibitor developed as a
female-controlled topical pre-exposure prophylaxis against HIV.

Vaginal dosing inverts the usual PBPK question. The therapeutic target is
*local*: the model must predict high concentrations in cervicovaginal
fluid and vaginal tissue while keeping systemic exposure low enough to
avoid adverse events. That needs an explicit description of drug release
from the dosage form into the luminal fluid, of transfer across the
vaginal epithelium and stroma, and of the pH gradient between the acidic
lumen (pH 4.4) and the neutral tissue (pH 7.0), which for a drug with a
pKa of 5.8 changes the ionised fraction by nearly three orders of
magnitude across a single cell layer.

The paper carries **one** structural model applied to two dosage forms:

* the **Ring-004 intravaginal ring**, a 25 mg dapivirine load in a
  platinum-catalysed silicone elastomer, worn for 21-35 days at a time;
* a **quick-dissolve vaginal film** with a 1.25 mg load and a 10 min
  dissolution half-life.

The two differ only in the drug-release law and in the
luminal-fluid-to-epithelium transfer rate constant, so they are extracted
as a single model file selected by the binary covariate `FORM_DPV_FILM`
(see the *Errata* section for why the release law had to be read across
four places in the paper).

```{r load-model}
mod <- readModelDb("Kay_2018_dapivirine_pbpk")
ui <- rxode2::rxode(mod)
ui$state
```

# Population

```{r population}
pop <- ui$population
str(pop)
```

The model was **not fitted to individual data**. It was assembled from
literature physiology for a typical 70 kg woman at rest (Table 1 of the
paper) and then calibrated so that its simulated profiles reproduced
published summary profiles from three clinical studies:

| Role | Study | Subjects |
|---|---|---|
| Model building, ring | Nel et al., reference [21] of the paper | 48 women, two multi-ring regimens |
| Model building, film | FAME 02 phase I, reference [30] | 15 women, seven daily films |
| External validation, ring | Nel et al., reference [23] | 16 women, one ring for 28 days |

Several physiological quantities had to be inferred rather than measured.
The vaginal epithelium volume is built up from a mean vaginal cell
diameter, the 87.46 cm^2 mean surface area of 62 vinyl polysiloxane
vaginal casts, the number of cell layers and a single-layer thickness.
The stroma volume is the epithelium volume times 14, from the observation
that the stroma is roughly 14 times thicker. Vaginal stromal blood flow
comes from nine women with *normal pregnancies*, which the authors flag
as likely an overestimate for the non-pregnant women being simulated.

# Source trace

Every value in `ini()` and every equation in `model()` comes from one of
the locations below. The equation set lives in the Supporting Information
file `BCP-84-1950-s004.docx` (Equations S1-S19); the main text prints
Equations 1-9, a subset.

| Model element | Source location |
|---|---|
| Luminal fluid, un-ionised and ionised (`vaginal_lumen`, `ion_vaginal_lumen`) | Supplement Eq. S1, S2 (= main text Eq. 1, 2) |
| Epithelium, un-ionised and ionised | Supplement Eq. S3, S4 |
| Stroma tissue, un-ionised and ionised | Supplement Eq. S5, S6 |
| Stromal blood (`vaginal_stroma_blood`) | Supplement Eq. S7 |
| Venous and arterial blood | Supplement Eq. S8, S9 (= main text Eq. 3, 4) |
| Lung, liver, rest-of-body | Supplement Eq. S10-S12 (= main text Eq. 5) |
| Hepatic elimination term | Supplement Eq. S13; main text Eq. 5 addendum |
| Ring release, per-ring decay clock | Supplement Eq. S14-S18; main text Eq. 6 and Eq. 9 |
| Film release | Supplement Eq. S19; main text Eq. 7 |
| Ionisation ratios (0.04 lumen, 15.85 tissue) | Main text Eq. 8, Henderson-Hasselbalch at pKa 5.8 |
| All compartment volumes, blood flows | Table 1 |
| All rate constants, partition coefficients, F, CL | Table 2 |
| Ring regimen timings (groups A and B) | Methods, *Ring simulations*; Supplement Eq. S14-S18 |
| Reproduction targets | Table 3 (concentrations), Table 4 (AUC), Table 5 (external validation), Figure 4 (film) |

A helper for the solves. The model is multi-output over thirteen ODE
states, so every `rxSolve()` call passes `useLinCmt = FALSE`; rxode2's
default ODE-to-`linCmt()` auto-conversion mis-maps multi-state models.
There is no inter-individual variability to suppress, so `omega` is never
touched.

```{r solve-helper}
solve_mod <- function(events) {
  out <- rxode2::rxSolve(
    mod, events,
    useLinCmt = FALSE,
    atol = 1e-12, rtol = 1e-10,
    returnType = "data.frame"
  )
  if (is.null(out$id)) out$id <- 1L
  out
}

# Observation rows always name an ODE STATE, never an algebraic observable:
# rxode2 returns Cc / Cfluid / Cepithelium / Cstroma as columns regardless.
obs_grid <- function(times, form) {
  data.frame(
    id = 1L, time = times, amt = NA_real_, evid = 0L,
    cmt = "venous", FORM_DPV_FILM = form
  )
}

# A ring insertion replaces whatever is left of the previous ring (evid = 5),
# which also restarts the tad(depot) decay clock that Eq. S14-S18 require.
ring_event <- function(time, amt) {
  data.frame(
    id = 1L, time = time, amt = amt, evid = 5L,
    cmt = "depot", FORM_DPV_FILM = 0
  )
}

trapz <- function(t, y) sum(diff(t) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)
```

# Ring, dosing regimen A

Group A of the model-building study wore two rings over 56 days: ring 1
for days 0-28, a three-day ring-free interval, then ring 2 for days
31-59. Simulations run to 100 days so the washout is captured.

```{r ring-A}
grid_A <- seq(0, 2400, by = 1)
ev_A <- bind_rows(
  ring_event(0, 25),
  ring_event(672, 0), # ring 1 removed at day 28
  ring_event(744, 25), # ring 2 inserted at day 31
  ring_event(1416, 0), # ring 2 removed at day 59
  obs_grid(grid_A, form = 0)
) |>
  arrange(time, dplyr::desc(evid))

sim_A <- solve_mod(ev_A)
range(sim_A$Cfluid)
```

```{r fig-ring-A, fig.cap = "Replicates Figure 2 of Kay 2018: dapivirine in (A) vaginal fluid, (B) plasma, (C) epithelium and (D) stroma over two consecutive rings."}
long_A <- sim_A |>
  select(time, Cfluid, Cc, Cepithelium, Cstroma) |>
  pivot_longer(-time, names_to = "matrix", values_to = "conc") |>
  mutate(matrix = factor(
    matrix,
    levels = c("Cfluid", "Cc", "Cepithelium", "Cstroma"),
    labels = c(
      "A. Vaginal luminal fluid", "B. Plasma",
      "C. Vaginal epithelium", "D. Vaginal stroma"
    )
  ))

ggplot(long_A, aes(time / 24, conc)) +
  geom_line() +
  facet_wrap(~matrix, scales = "free_y") +
  labs(x = "Time (days)", y = "Dapivirine concentration (mg/L)") +
  theme_bw()
```

The paper tabulates its own single-patient simulation at five landmarks
(Table 3A). Because the model is deterministic -- no random effects, no
residual error -- these are exact reproduction targets rather than
statistical comparisons, so the assertions below are tight.

```{r ring-A-landmarks}
at <- function(sim, tt, col) sim[[col]][which.min(abs(sim$time - tt))]

landmarks_A <- tibble::tribble(
  ~Landmark, ~Matrix, ~Reported, ~Simulated,
  "C prior to ring 1 removal (672 h)", "Vaginal fluid", 28.96, at(sim_A, 672, "Cfluid"),
  "C at insertion of ring 2 (744 h)", "Vaginal fluid", 0.53, at(sim_A, 744, "Cfluid"),
  "C prior to ring 2 removal (1416 h)", "Vaginal fluid", 28.96, at(sim_A, 1416, "Cfluid"),
  "C prior to ring 1 removal (672 h)", "Plasma", 2.48e-4, at(sim_A, 672, "Cc"),
  "C at insertion of ring 2 (744 h)", "Plasma", 1.65e-4, at(sim_A, 744, "Cc"),
  "C prior to ring 2 removal (1416 h)", "Plasma", 2.49e-4, at(sim_A, 1416, "Cc")
) |>
  mutate(`% diff` = 100 * (Simulated - Reported) / Reported)

knitr::kable(landmarks_A, digits = c(0, 0, 5, 5, 2))

# Deterministic solve against the paper's own deterministic simulation: the
# only sources of difference are the paper's 1 h output grid and its rounding
# to 4 significant figures. 2% is generous for that and still breaks on any
# mis-transcribed rate constant or volume.
stopifnot(max(abs(landmarks_A$`% diff`)) < 2)
```

# Ring, dosing regimen B

Group B wore three rings over 57 days: ring 1 for days 0-35, a three-day
break, ring 2 for days 38-59, then ring 3 immediately for 24 h. Each
insertion restarts that ring's own release-decay clock, which is what
makes ring 3 -- worn for a single day -- produce the highest luminal
concentration of the regimen.

```{r ring-B}
ev_B <- bind_rows(
  ring_event(0, 25),
  ring_event(840, 0), # ring 1 removed at day 35
  ring_event(912, 25), # ring 2 inserted at day 38
  ring_event(1416, 25), # ring 3 inserted at day 59, immediately on removal of ring 2
  ring_event(1440, 0), # ring 3 removed at day 60
  obs_grid(seq(0, 2400, by = 1), form = 0)
) |>
  arrange(time, dplyr::desc(evid))

sim_B <- solve_mod(ev_B)

landmarks_B <- tibble::tribble(
  ~Landmark, ~Matrix, ~Reported, ~Simulated,
  "C prior to ring 1 removal (840 h)", "Vaginal fluid", 23.56, at(sim_B, 840, "Cfluid"),
  "C at insertion of ring 2 (912 h)", "Vaginal fluid", 0.43, at(sim_B, 912, "Cfluid"),
  "C prior to ring 2 removal (1416 h)", "Vaginal fluid", 35.65, at(sim_B, 1416, "Cfluid"),
  "C prior to ring 3 removal (1440 h)", "Vaginal fluid", 56.88, at(sim_B, 1440, "Cfluid"),
  "C prior to ring 1 removal (840 h)", "Plasma", 2.00e-4, at(sim_B, 840, "Cc"),
  "C at insertion of ring 2 (912 h)", "Plasma", 1.35e-4, at(sim_B, 912, "Cc"),
  "C prior to ring 2 removal (1416 h)", "Plasma", 2.98e-4, at(sim_B, 1416, "Cc"),
  "C prior to ring 3 removal (1440 h)", "Plasma", 3.06e-4, at(sim_B, 1440, "Cc")
) |>
  mutate(`% diff` = 100 * (Simulated - Reported) / Reported)

knitr::kable(landmarks_B, digits = c(0, 0, 5, 5, 2))

stopifnot(max(abs(landmarks_B$`% diff`)) < 3)
```

```{r fig-ring-B, fig.cap = "Replicates Figure 3 of Kay 2018: three consecutive rings, group B regimen."}
sim_B |>
  select(time, Cfluid, Cc, Cepithelium, Cstroma) |>
  pivot_longer(-time, names_to = "matrix", values_to = "conc") |>
  mutate(matrix = factor(
    matrix,
    levels = c("Cfluid", "Cc", "Cepithelium", "Cstroma"),
    labels = c(
      "A. Vaginal luminal fluid", "B. Plasma",
      "C. Vaginal epithelium", "D. Vaginal stroma"
    )
  )) |>
  ggplot(aes(time / 24, conc)) +
  geom_line() +
  facet_wrap(~matrix, scales = "free_y") +
  labs(x = "Time (days)", y = "Dapivirine concentration (mg/L)") +
  theme_bw()
```

# Film

A single 1.25 mg quick-dissolve film. The bioavailability assumption
enters here as a dose reduction (`f(depot)` = 0.30, so 0.375 mg is
available); for the ring the same 30% is already inside the release rate
constant, so the ring dose is the full 25 mg load.

```{r film}
ev_F <- bind_rows(
  data.frame(
    id = 1L, time = 0, amt = 1.25, evid = 1L,
    cmt = "depot", FORM_DPV_FILM = 1
  ),
  obs_grid(seq(0, 200, by = 0.05), form = 1)
) |>
  arrange(time, dplyr::desc(evid))

sim_F <- solve_mod(ev_F)

film_landmarks <- tibble::tribble(
  ~Quantity, ~Source, ~Reported, ~Simulated,
  "Plasma Cmax (mg/L)", "Figure 4B", 7.0e-4, max(sim_F$Cc),
  "Plasma Tmax (h)", "Figure 4B", 7.5, sim_F$time[which.max(sim_F$Cc)],
  "Stroma Cmax (mg/L)", "Figure 4D", 0.12, max(sim_F$Cstroma),
  "Stroma Tmax (h)", "Results, film", 3.0, sim_F$time[which.max(sim_F$Cstroma)],
  "Epithelium Tmax (h)", "Results, film", 1.0, sim_F$time[which.max(sim_F$Cepithelium)],
  "Luminal fluid Tmax (h)", "Results, film", 0.5, sim_F$time[which.max(sim_F$Cfluid)]
)
knitr::kable(film_landmarks, digits = 5)
```

Two claims in the film Results are checked as statements rather than
numbers. The simulated epithelial concentration must sit inside the
1.12-4.75 mg/L band the paper reports for its epithelium visual
predictive check, and the plasma concentration must stay above the HIV
IC50 (0.09-0.14 ng/mL) for roughly 200 h.

```{r film-claims}
epi_cmax <- max(sim_F$Cepithelium)
above_ic50 <- max(sim_F$time[sim_F$Cc > 0.09e-3])

stopifnot(
  # Paper's VPC epithelium range, Results section 'Film VPC'.
  epi_cmax > 1.12, epi_cmax < 4.75,
  # "remains ... above the HIV IC50 ... for approximately 200 h".
  above_ic50 >= 180
)
c(epithelium_cmax = epi_cmax, hours_above_ic50 = above_ic50)
```

```{r fig-film, fig.cap = "Replicates Figure 4 of Kay 2018: a single vaginal film over the first 26 h."}
sim_F |>
  filter(time <= 26) |>
  select(time, Cfluid, Cc, Cepithelium, Cstroma) |>
  pivot_longer(-time, names_to = "matrix", values_to = "conc") |>
  mutate(matrix = factor(
    matrix,
    levels = c("Cfluid", "Cc", "Cepithelium", "Cstroma"),
    labels = c(
      "A. Vaginal luminal fluid", "B. Plasma",
      "C. Vaginal epithelium", "D. Vaginal stroma"
    )
  )) |>
  ggplot(aes(time, conc)) +
  geom_line() +
  facet_wrap(~matrix, scales = "free_y") +
  labs(x = "Time (h)", y = "Dapivirine concentration (mg/L)") +
  theme_bw()
```

# External validation: a single 28-day ring

Table 5 of the paper reports the model's prediction for a study that took
no part in model building -- 16 women wearing one Ring-004 for 28 days.

```{r ring-single}
ev_S <- bind_rows(
  ring_event(0, 25),
  obs_grid(seq(0, 840, by = 0.25), form = 0)
) |>
  arrange(time, dplyr::desc(evid))

sim_S <- solve_mod(ev_S)
win_S <- filter(sim_S, time <= 672)

tab5 <- tibble::tribble(
  ~Quantity, ~Reported, ~Simulated,
  "Vaginal fluid C at 1.5 h (mg/L)", 4.75, at(win_S, 1.5, "Cfluid"),
  "Vaginal fluid C prior to ring removal (mg/L)", 31.20, at(win_S, 672, "Cfluid"),
  "Plasma C prior to ring removal (mg/L)", 2.65e-4, at(win_S, 672, "Cc")
) |>
  mutate(`% diff` = 100 * (Simulated - Reported) / Reported)
knitr::kable(tab5, digits = c(0, 5, 5, 1))
```

These three sit about 10% from the paper, further than the Table 3
landmarks did, so they are reported rather than asserted. The discrepancy
is inside the paper rather than in this encoding: Table 3A reports the
*same* quantity under the *same* 28-day ring as plasma 2.48e-4 mg/L and
vaginal fluid 28.96 mg/L, and those two are reproduced to three
significant figures above. The Table 5 column was evidently read at a
slightly different time point or from a slightly different run.

```{r ring-single-consistency}
# The Table 3A landmark is reproduced from this single-ring simulation too,
# which is what localises the discrepancy to the paper's Table 5 column.
c(
  fluid_672 = at(win_S, 672, "Cfluid"), # Table 3A: 28.96
  plasma_672 = at(win_S, 672, "Cc") # Table 3A: 2.48e-4
)
```

# PKNCA validation

The NCA is run with PKNCA over each group's **first** ring-wear period --
0-28 days for group A and 0-35 days for group B -- because that is the
window the paper's own AUC table (Table 4) and its Cmax / Tmax columns
(Table 3) use. Both matrices the paper tabulates are carried through. The
concentration frame is filtered only on `!is.na()`, so the time-zero
record survives and PKNCA has a defined AUC start.

```{r pknca}
nca_frame <- function(sim, end_h, grp) {
  w <- filter(sim, time <= end_h)
  bind_rows(
    transmute(w, id, time, Cc = Cfluid, matrix = "Vaginal fluid"),
    transmute(w, id, time, Cc, matrix = "Plasma")
  ) |>
    mutate(group = grp)
}

nca_input <- bind_rows(
  nca_frame(sim_A, 672, "Group A"),
  nca_frame(sim_B, 840, "Group B")
) |>
  filter(!is.na(Cc))

stopifnot(
  nrow(nca_input) > 0,
  all(nca_input$Cc >= 0),
  # Four group-by-matrix combinations, each with a time-zero record.
  nrow(distinct(nca_input, group, matrix)) == 4L,
  sum(nca_input$time == 0) == 4L
)

o_conc <- PKNCA::PKNCAconc(
  nca_input, Cc ~ time | group + matrix + id,
  concu = "mg/L", timeu = "h"
)
dose_df <- nca_input |>
  distinct(group, matrix, id) |>
  mutate(time = 0, amt = 25)
o_dose <- PKNCA::PKNCAdose(
  dose_df, amt ~ time | group + matrix + id,
  doseu = "mg"
)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE
)

res <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals))
nca <- as.data.frame(res) |> filter(!is.na(PPORRES))
knitr::kable(
  nca |>
    select(group, matrix, PPTESTCD, PPORRES) |>
    rename(
      "Regimen" = group, "Matrix" = matrix,
      "Parameter" = PPTESTCD, "Value" = PPORRES
    ),
  digits = 6
)
```

## Comparison against the published NCA

The reference column below is the paper's **simulated** column, not its
observed column: the object under test is whether this nlmixr2 encoding
reproduces the authors' own MATLAB SimBiology run. AUC comes from Table 4
(AUC0-28 for group A, AUC0-35 for group B) and Cmax / Tmax from Table 3.

```{r nca-comparison}
sim_nca <- nca |>
  filter(PPTESTCD %in% c("auclast", "cmax", "tmax")) |>
  select(group, matrix, PPTESTCD, PPORRES)

ref_nca <- data.frame(
  group = c("Group A", "Group A", "Group B", "Group B"),
  matrix = c("Vaginal fluid", "Plasma", "Vaginal fluid", "Plasma"),
  auclast = c(29079, 0.177, 33067, 0.211), # Table 4, Simulated column
  cmax = c(60.06, 3.30e-4, 60.06, 3.30e-4), # Table 3, Simulated column
  tmax = c(71.51, 315.33, 69.67, 317.86) # Table 3, Simulated column
)

cmp <- ncaComparisonTable(
  sim_nca, ref_nca,
  by = c("group", "matrix"),
  units = c(auclast = "mg*h/L", cmax = "mg/L", tmax = "h")
)
knitr::kable(cmp, digits = 5)
attr(cmp, "footnote")
```

```{r nca-assert}
pct <- suppressWarnings(as.numeric(gsub("[^0-9.eE+-]", "", cmp$`% diff`)))
stopifnot(length(pct) == nrow(cmp), !anyNA(pct))

# Deterministic reproduction of a deterministic source simulation, so a tight
# bound is the right gate here -- there is no cohort draw to vary across
# machines. The residual differences are the paper's 1 h output grid (which
# moves Tmax by up to 4% on a very flat maximum) and its 4-significant-figure
# rounding; realised max was 4.0%. A mis-transcribed rate constant, volume or
# blood flow moves these by tens of percent.
stopifnot(max(abs(pct)) < 6)
max(abs(pct))
```

# Parameter-uncertainty cohort

The paper's visual predictive checks sampled the coefficients of
variation in Tables 1 and 2 (1000 replicates of 30 women). The version
below draws a single 200-woman cohort, which is enough to show the
spread; it is not a reproduction of the paper's centile-of-centile
construction.

Because the model carries no estimated random effects, the variability is
introduced by resampling the parameters themselves, exactly as the source
did: log-normal on the rate constants, clearance and partition
coefficients, normal on body weight, and uniform 25-35% on
bioavailability.

```{r vpc-cohort}
rxode2::rxSetSeed(20180802)
set.seed(20180802)
n_sub <- 200L

ln_draw <- function(n, mu, cv) mu * exp(rnorm(n, 0, sqrt(log(cv^2 + 1))))

cohort <- data.frame(
  id = seq_len(n_sub),
  v_lumen = ln_draw(n_sub, 0.0005, 0.71), # Table 1, CV 71%
  q_cardiac = ln_draw(n_sub, 294.0, 0.30), # Table 1, CV 30%
  k_lumen_epi_ring = ln_draw(n_sub, 0.058, 0.30), # Table 2, CV 30%
  k_epi_stroma_active = ln_draw(n_sub, 0.5, 0.30),
  k_stroma_blood = ln_draw(n_sub, 0.122, 0.30),
  k_stroma_blood_active = ln_draw(n_sub, 0.5, 0.30),
  k_blood_stroma = ln_draw(n_sub, 0.01, 0.30),
  k_ring_decay = ln_draw(n_sub, 1.2e-3, 0.10), # Table 2, CV 10%
  lcl = log(ln_draw(n_sub, 4, 0.10)) # Table 2, CV 10%
)
# Blood flows keep their typical-patient ratio to cardiac output (Table 1
# footnote d).
cohort$q_stroma <- 14.3 / 294.0 * cohort$q_cardiac
cohort$q_liver <- 98.8 / 294.0 * cohort$q_cardiac
cohort$q_other <- 180.9 / 294.0 * cohort$q_cardiac

ev_vpc <- bind_rows(
  cohort |> transmute(id, time = 0, amt = 25, evid = 5L, cmt = "depot", FORM_DPV_FILM = 0),
  tidyr::crossing(
    cohort |> select(id),
    time = seq(0, 840, by = 6)
  ) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "venous", FORM_DPV_FILM = 0)
) |>
  arrange(id, time, dplyr::desc(evid))

sim_vpc <- rxode2::rxSolve(
  mod, ev_vpc, params = cohort,
  useLinCmt = FALSE, atol = 1e-10, rtol = 1e-8,
  returnType = "data.frame"
)
```

```{r fig-vpc, fig.cap = "Parameter-uncertainty cohort (200 women), luminal fluid and plasma; compare Figures 5 and 8 of Kay 2018. Band is the 5th-95th percentile, line the median."}
vpc_summary <- sim_vpc |>
  select(time, Cfluid, Cc) |>
  pivot_longer(-time, names_to = "matrix", values_to = "conc") |>
  mutate(matrix = factor(matrix,
    levels = c("Cfluid", "Cc"),
    labels = c("Vaginal luminal fluid", "Plasma")
  )) |>
  group_by(matrix, time) |>
  summarise(
    lo = quantile(conc, 0.05), mid = median(conc), hi = quantile(conc, 0.95),
    .groups = "drop"
  )

ggplot(vpc_summary, aes(time / 24, mid)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.25) +
  geom_line() +
  facet_wrap(~matrix, scales = "free_y") +
  labs(x = "Time (days)", y = "Dapivirine concentration (mg/L)") +
  theme_bw()
```

```{r vpc-assert}
peak <- vpc_summary |>
  filter(matrix == "Vaginal luminal fluid", time <= 672) |>
  summarise(med = max(mid), lo = max(lo), hi = max(hi))

# The typical-value luminal Cmax is 60 mg/L, and the cohort median must stay
# near it: a resampling scheme that perturbed the central tendency rather than
# the spread would move this. Robust statistics only -- the 5th and 95th
# percentiles of one 200-subject draw are not reproducible across rxode2
# builds or thread counts (see the repo note on cohort assertions), so the
# band is checked only for being a band.
stopifnot(
  abs(peak$med - 60.08) / 60.08 < 0.25,
  peak$lo < peak$med,
  peak$med < peak$hi
)
peak
```

# Errata and assumptions

**The ring release exponent is printed without its minus sign in two of
the four places the paper states it.** Main-text Equation 6 and
Supplement Equation S14 both typeset
`dD_Ring/dt = -(ka * exp(ka_exp * (t - t_Ring))) * D_Ring`, a release rate
that *grows* 2.2-fold over 28 days. Three independent lines of evidence
say the exponent is negative and the model here uses the decaying form:

1. Main-text **Equation 9** prints it with the minus sign,
   `ka(t) = ka * exp(-ka_exp)`.
2. The surrounding prose is explicit: "we allow the initial rate of drug
   release (ka) to decay exponentially over time according to the rate
   constant, ka_exp", motivated by a depletion zone forming at the ring
   surface.
3. Only the decaying form reproduces the paper's own reported simulated
   exposures. Integrating the release over 28 days gives 0.83 mg released
   with the negative exponent and 1.82 mg with the positive one; the
   resulting luminal AUC0-28 is about 29,000 mg*h/L versus about 63,000,
   against a reported simulated value of 29,079.

Equation 9 also drops the `* t` from its exponent, and Equation 6
attaches the `(t - t_Ring)` that Equation 9 omits; the two together give
the intended `exp(-ka_exp * (t - t_Ring))`.

**Table 2 transposes the Description labels of the two
luminal-fluid-to-epithelium rate constants against their Abbreviation
column.** The row reading "Ring: luminal fluids to epithelium" carries
the abbreviation `kL,E_film` and the value 2.773, and the row reading
"Film: ..." carries `kL,E_ring` and 0.058. The abbreviations are correct:
0.058 is cited to reference [23], the 12 h luminal half-life measured
after **ring** removal (ln(2)/12 = 0.0578), while the text derives the
**film** value as roughly 50-fold larger. The model uses
`k_lumen_epi_ring` = 0.058 and `k_lumen_epi_film` = 2.773, and this
reproduces the ring results exactly.

**The stated film plasma Cmax misreads the paper's own figure.** The
Results text says the simulated plasma concentration "gradually increases
to a Cmax of 0.0012 mg l-1 approximately 10 h after film insertion".
Panel B of Figure 4 shows the simulated curve plateauing at about
0.0007 mg/L around 7-8 h, on an axis gridded at 0.0002 intervals up to
0.0014. This encoding gives 7.07e-4 mg/L at 7.75 h, matching the figure;
the prose value is not reproduced and is treated as a reading error.
Panel D of the same figure independently confirms the encoding: its
stroma peak of about 0.12 mg/L near 3 h is reproduced as 0.1234 mg/L at
2.89 h.

**State variables are concentrations, not amounts.** The main text says
"D is the amount of drug (mg)", but every supplement equation has the
form `(1/V) * [rate * D * V]` and the systemic equations divide by
partition coefficients, both of which are only dimensionally coherent if
`D` is a concentration. The supplement's own section headings agree
("Unionised and ionised drug **concentration** in the vaginal
epithelium"). The dosage-form depot is the one exception and is carried
as an amount in mg, as in the source.

**The film volume `V_film` (9.032e-5 L, Table 1) is not carried as a
parameter.** Supplement Equation S19 writes the film state as a
concentration and multiplies by `V_film` on the way into the lumen, so
the volume cancels exactly. Encoding the film state as an amount in mg
gives an identical system with one fewer parameter and lets the dose
record carry the real 1.25 mg film load.

**Bioavailability is applied asymmetrically, as in the source.** The 30%
assumption is back-solved into the ring's release rate constant
(`ka = -ln(23.8/25)/28/24`), so the ring's `f(depot)` is 1 and its dose is
the full 25 mg load. For the film it is a dose reduction, `f(depot)` =
0.30 on a 1.25 mg load. Reading it as a rate reduction on `kF,L` instead
puts the simulated epithelial Cmax at 8.75 mg/L, well above the
1.12-4.75 mg/L band the paper reports for that quantity.

**Table 5 landmarks are reproduced less closely than Table 3
landmarks.** Table 5's single-ring "C prior to ring removal" is
2.65e-4 mg/L (plasma) and 31.20 mg/L (vaginal fluid), against 2.48e-4 and
28.96 for the same quantity under the same 28-day ring in Table 3A. This
encoding reproduces the Table 3A pair to three significant figures, which
places the discrepancy inside the paper rather than in the encoding --
most likely a different sampling time behind the Table 5 column.

**No inter-individual variability or residual error is encoded.** The
source is a deterministic typical-value model; population variability was
produced by Monte Carlo resampling of the Table 1 and Table 2
coefficients of variation, which the *Parameter-uncertainty cohort*
section above reproduces at the simulation level rather than as `eta`
terms. Consequently `rxSolve()` is never called with `omega`.

**Compartment naming.** The vaginal luminal fluid, epithelium, stroma
tissue and stromal blood, and the un-ionised/ionised species split, have
no canonical compartment names in the library -- this is its first
vaginal-route model -- so they are declared through
`paper_specific_compartments`. The specimen vocabulary likewise has no
vaginal-fluid entry; `epithelial lining fluid` is used for the luminal
compartments as the closest available term for a fluid lining an
epithelium. The vascular pools are recorded as `plasma` because every
partition coefficient in Table 2 is a tissue-to-plasma ratio and the
source compares the venous state directly against measured plasma.

**Physiological assumptions carried from the source, unchanged.** Vaginal
stromal blood flow is taken from pregnant women and is likely an
overestimate; the stromal blood volume (0.108 L) is a dermal-blood-volume
proxy chosen to replicate trial results; the partition coefficients come
from mice given intravaginal dapivirine nanoparticles, with the lung value
derived from an assumed below-limit-of-detection concentration and the
rest-of-body value calibrated rather than measured; and the stroma volume
is the epithelium volume times 14. The paper discusses each of these as a
limitation, and its own conclusion -- that simulated stromal
concentrations are about 14-fold below epithelial ones purely because of
that volume assumption -- is reproduced here.
