---
title: "Nemtabrutinib (Kemal 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nemtabrutinib (Kemal 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Kemal_2026_nemtabrutinib"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Kemal_2026_nemtabrutinib"))$description`
* Article: <https://doi.org/10.1002/psp4.70257>

This paper contributes four models to the library. The population PK
model is the subject of the first half of this vignette; the three
logistic exposure-response models it feeds are covered in the
"Exposure-response models" section below.

| Model | What it describes |
|-------|-------------------|
| `Kemal_2026_nemtabrutinib` | Two-compartment population PK, full covariate model |
| `Kemal_2026_nemtabrutinib_bor` | Exposure-efficacy: best overall response in CLL/SLL |
| `Kemal_2026_nemtabrutinib_ae` | Exposure-safety: any-grade drug-related adverse events |
| `Kemal_2026_nemtabrutinib_hypertension` | Exposure-safety: any-grade hypertension |

# Population

The pooled analysis population comprised 578 adults with hematologic
malignancies enrolled in two Merck-sponsored trials of the oral
non-covalent BTK inhibitor nemtabrutinib (also known as MK-1026 and
formerly ARQ 531):

* BELLWAVE-001 (NCT03162536), a phase 1/2 dose-escalation study
  (n = 136, doses 5-75 mg qd), enrolled patients with relapsed /
  refractory CLL/SLL, B-cell non-Hodgkin lymphoma, and Waldenstrom's
  macroglobulinemia.
* BELLWAVE-003 (NCT04728893), a phase 2 dose-escalation and
  confirmation study (n = 442, doses 45 / 65 / 80 mg qd), enrolled
  patients with CLL/SLL, marginal zone / follicular / mantle-cell
  lymphomas, Richter's transformation, and Waldenstrom's
  macroglobulinemia.

Baseline demographics (Table 1 of Kemal et al. 2026): median age 68
years (range 25-89), median body weight 74 kg (range 41-147),
median baseline albumin 4.14 g/dL (i.e. 41.4 g/L), 34.1% female,
85.8% White, 7.8% Asian, 2.4% Black, 2.4% Other, 1.6% Missing.
Primary diagnosis: 49.8% CLL/SLL, 30.6% other hematologic
malignancies, 9.7% B-cell non-Hodgkin lymphoma, 9.7% Waldenstrom's
macroglobulinemia. Renal function: 29.2% normal, 38.6% mild
impairment, 30.1% moderate impairment; hepatic function: 80.3%
normal, 18.3% mild impairment, 1.4% moderate impairment. Concomitant
medication exposure at any observation: 27.7% weak / 3.3% moderate /
2.0% strong CYP3A4 inhibitor; 46.6% weak / 2.8% moderate CYP3A4
inducer; 25.6% PPI; 6.8% H2 antagonist; 10.8% antacid (Table S3).
Model development used 5669 non-BLQ observations.

The same information is available programmatically via
`readModelDb("Kemal_2026_nemtabrutinib")$population`.

# Source trace

Every `ini()` value in
`inst/modeldb/specificDrugs/Kemal_2026_nemtabrutinib.R` carries an
in-file comment identifying its source location; the table below
collects the audit trail in one place. All final estimates come from
Table 2 of the main article; the reference values for continuous
covariates and the exact form of the multiplicative covariate
composition come from the NONMEM control stream in the supplement.

| Component | Value | Source |
|-----------|-------|--------|
| Two-compartment PK with first-order absorption + lag, first-order elimination | -- | Kemal 2026 Methods 2.2.1 and NONMEM `$SUBROUTINES ADVAN4 TRANS4` |
| CL/F (typical) | 3.33 L/h | Table 2 |
| Vc/F (typical) | 120 L | Table 2 |
| Ka | 2.83 1/h | Table 2 |
| Q/F | 0.681 L/h | Table 2 |
| Vp/F | 66.9 L | Table 2 |
| Absorption lag | 0.494 h | Table 2 |
| Weight exponent on CL (fixed) | 0.331 | Table 2 (fixed after WT-only fit; Methods 2.2.2) |
| Weight exponent on Vc (fixed) | 0.807 | Table 2 (fixed after WT-only fit; Methods 2.2.2) |
| Reference weight | 73.5 kg | NONMEM supplement `$PK` (`WT = 73.5` if missing) |
| Reference age | 68 years | NONMEM supplement `$PK` (`AGEM = 68` if missing) |
| Reference albumin | 41.2 g/L | NONMEM supplement `$PK` (`BALBM = 41.2` if missing) |
| Age power exponent on CL | -0.503 | Table 2 |
| Albumin power exponent on CL | -0.395 | Table 2 |
| Sex effect on CL (female vs male ref) | -0.133 | Table 2 |
| Sex effect on Vc | -0.0901 | Table 2 |
| Race effect on CL (Black vs White ref) | 0.0533 | Table 2 |
| Race effect on CL (Asian vs White ref) | -0.117 | Table 2 |
| Race effect on Vc (Black) | 0.102 | Table 2 |
| Race effect on Vc (Asian) | -0.116 | Table 2 |
| Disease effect on CL (B-cell NHL vs CLL/SLL ref) | -0.166 | Table 2 |
| Disease effect on CL (WM vs CLL/SLL) | 0.0718 | Table 2 |
| Disease effect on CL (Other vs CLL/SLL) | -0.0244 | Table 2 |
| Disease effect on Vc (B-cell NHL) | 0.00224 | Table 2 |
| Disease effect on Vc (WM) | 0.152 | Table 2 |
| Disease effect on Vc (Other) | -0.0200 | Table 2 |
| Mild renal impairment effect on CL | 0.0537 | Table 2 |
| Moderate renal impairment effect on CL | -0.00186 | Table 2 |
| Mild hepatic impairment effect on CL | 0.00401 | Table 2 |
| Moderate CYP3A4 inducer effect on CL | 0.0220 | Table 2 |
| Strong CYP3A4 inhibitor effect on CL | -0.0119 | Table 2 |
| Low-dose (<30 mg) effect on F | -0.151 | Table 2 |
| PPI effect on F | 0.00232 | Table 2 |
| H2 antagonist effect on F | 0.0264 | Table 2 |
| Antacid effect on F | -0.0360 | Table 2 |
| CL IIV | 39.8% CV -> omega^2 = 0.14710 | Table 2 (log(1 + 0.398^2)) |
| Vc IIV | 17.1% CV -> omega^2 = 0.02884 | Table 2 (log(1 + 0.171^2)) |
| Proportional residual SD | 0.222 | Table 2 |
| Additive residual SD | 3.19 ng/mL | Table 2 |
| S2 = V2/1000 -> Cc = 1000 * central / vc | -- | NONMEM `$PK S2 = V2 / 1000` (dose in mg, Vc in L, Cc in ng/mL) |
| Terminal half-life derived from parameter estimates | 85 h | Kemal 2026 Section 3.2 |
| Tmax (observed median) | ~2 h | Kemal 2026 Introduction |
| Accumulation ratio (AUC0-24) | 1.8 to 3.4 | Kemal 2026 Introduction |
| Accumulation ratio (Cmax) | 1.53 to 2.85 | Kemal 2026 Introduction |

# Virtual cohort

Original observed data are not publicly available. The figures below
use small virtual populations (n = 50 per dose arm across the three
BELLWAVE-003 dose levels of 45, 65, and 80 mg qd; 150 subjects total,
well under the 200 per arm cap) whose covariate distributions
approximate the pooled analysis population (Kemal 2026 Table 1). All
covariates are set to the median / reference-category typical
individual to keep the simulation deterministic-typical; between-arm
differences reflect dose only. Reproducing the paper's forest-plot
covariate simulations (Figure 1) would require the full covariate
distributions and is out of scope for this per-model validation.

```{r cohort}
set.seed(19260517)

typical_covs <- list(
  WT                       = 73.5,   # reference weight (NONMEM $PK)
  AGE                      = 68,     # reference age
  ALB                      = 41.2,   # reference albumin, g/L
  SEXF                     = 0L,     # male reference
  RACE_ASIAN               = 0L,     # White reference
  RACE_BLACK               = 0L,
  DIS_BCELLNHL             = 0L,     # CLL/SLL reference
  DIS_WM                   = 0L,
  DIS_OTHER_HEME           = 0L,
  RENALIMP_MILD            = 0L,     # normal renal function
  RENALIMP_MOD             = 0L,
  HEPIMP_MILD              = 0L,     # normal hepatic function
  CONMED_CYP3A4_IND_MOD    = 0L,     # no comedication
  CONMED_CYP3A4_INH_STRONG = 0L,
  CONMED_PPI               = 0L,
  CONMED_H2RA              = 0L,
  CONMED_ANTACID           = 0L
)

make_cohort <- function(dose_mg, n_id, id_offset) {
  # 28 daily doses at 0, 24, ..., 24*27 h; dense observation grid over
  # first 24 h and around each trough, plus post-dose-28 half-life
  # window to expose the terminal phase.
  dose_times   <- seq(0, by = 24, length.out = 28)
  obs_times    <- sort(unique(c(
    seq(0, 24, by = 0.5),                        # first-dose PK profile
    dose_times,                                  # pre-dose troughs
    dose_times + 2,                              # near-Tmax at each dose
    seq(24 * 27, 24 * 27 + 24, by = 1),          # last-dose PK profile
    seq(24 * 28, 24 * 28 + 24 * 10, by = 6)      # terminal-phase tail
  )))

  cov_cols <- as.data.frame(typical_covs)

  ids <- id_offset + seq_len(n_id)

  doses <- tidyr::crossing(id = ids, time = dose_times) |>
    dplyr::mutate(evid = 1L, amt = dose_mg, cmt = "depot", DOSE = dose_mg)

  obs <- tidyr::crossing(id = ids, time = obs_times) |>
    dplyr::mutate(evid = 0L, amt = 0, cmt = "central", DOSE = dose_mg)

  ev <- dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    dplyr::mutate(treatment = paste0(dose_mg, " mg qd"))

  cbind(ev, cov_cols)
}

events <- dplyr::bind_rows(
  make_cohort(45, n_id = 50, id_offset =   0L),
  make_cohort(65, n_id = 50, id_offset = 100L),
  make_cohort(80, n_id = 50, id_offset = 200L)
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod <- rxode2::rxode2(readModelDb("Kemal_2026_nemtabrutinib"))

sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep   = c("treatment", "DOSE"),
  addDosing = FALSE
) |>
  as.data.frame()
```

# Replicate published patterns

The paper does not publish observed concentration-time data (the
BELLWAVE trials are ongoing) and does not publish a summary NCA
table. Instead the Introduction reports narrative descriptors
(median Tmax ~ 2 h, dose-proportional exposure at >= 30 mg,
accumulation ratios 1.8-3.4 for AUC0-24 and 1.53-2.85 for Cmax) and
Section 3.2 reports a derived terminal half-life of 85 h. Figure 2
plots per-subject Cavg and Cmax across dose levels but the underlying
values are not tabulated. The chunks below reproduce the exposure
patterns implied by the packaged full-covariate model at the three
BELLWAVE-003 doses (45, 65, 80 mg qd).

## First-dose PK profile

```{r first-dose}
sim |>
  dplyr::filter(time <= 24) |>
  ggplot(aes(time, Cc, group = interaction(id, treatment), colour = treatment)) +
  geom_line(alpha = 0.35) +
  facet_wrap(~treatment) +
  labs(
    x = "Time after first dose (h)",
    y = "Nemtabrutinib plasma concentration (ng/mL)",
    title = "First-dose PK profile by dose level",
    caption = "Per-subject simulations (n = 50 per arm, typical covariates)."
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

## Steady-state trough progression across the first 28 days

```{r trough}
dose_times_grid <- seq(0, by = 24, length.out = 28)

sim |>
  dplyr::filter(time %in% dose_times_grid, time > 0) |>
  dplyr::mutate(day = round(time / 24)) |>
  dplyr::group_by(treatment, day) |>
  dplyr::summarise(
    median_pre = median(Cc, na.rm = TRUE),
    p25_pre    = quantile(Cc, 0.25, na.rm = TRUE),
    p75_pre    = quantile(Cc, 0.75, na.rm = TRUE),
    .groups    = "drop"
  ) |>
  ggplot(aes(day, median_pre, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = p25_pre, ymax = p75_pre), alpha = 0.2, colour = NA) +
  geom_line(size = 0.7) +
  labs(
    x = "Study day (pre-dose)",
    y = "Nemtabrutinib pre-dose concentration (ng/mL)",
    title = "Pre-dose (trough) concentrations approaching steady state",
    caption = "Ribbons show IQR; lines show medians (n = 50 per arm). The paper reports 'plateau after ~15 days' (Introduction)."
  ) +
  theme_bw()
```

# PKNCA validation

Compute Cmax, Tmax, AUC0-24 on day 1, and derived terminal half-life
from the terminal-phase window after the 28th dose (last dose at t =
648 h). PKNCA formulas include the treatment grouping variable so
per-dose values are separately available.

```{r pknca-day1}
sim_day1 <- sim |>
  dplyr::filter(time <= 24, !is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time=0 row per (id, treatment); for extravascular pre-dose Cc=0.
sim_day1 <- dplyr::bind_rows(
  sim_day1,
  sim_day1 |> dplyr::distinct(id, treatment) |>
    dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

conc_day1 <- PKNCA::PKNCAconc(sim_day1, Cc ~ time | treatment + id)

dose_day1 <- events |>
  dplyr::filter(evid == 1, time == 0) |>
  dplyr::select(id, time, amt, treatment)

pk_dose_day1 <- PKNCA::PKNCAdose(dose_day1, amt ~ time | treatment + id)

intervals_day1 <- data.frame(
  start    = 0,
  end      = 24,
  cmax     = TRUE,
  tmax     = TRUE,
  auclast  = TRUE
)

nca_day1 <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_day1, pk_dose_day1, intervals = intervals_day1)
)
```

```{r pknca-terminal}
# Terminal-phase profile after the last (28th) dose: obs times
# 648 h ... 648 + 240 h. Use the paper's "85 h" claim as the check.
sim_terminal <- sim |>
  dplyr::filter(time >= 24 * 27, time <= 24 * 27 + 24 * 10, !is.na(Cc)) |>
  dplyr::mutate(time_after_last_dose = time - 24 * 27) |>
  dplyr::select(id, time = time_after_last_dose, Cc, treatment)

# Guarantee a time=0 row per (id, treatment) so PKNCA can anchor the
# terminal slope. Set Cc = 0 as the anchor -- half.life ignores the
# time=0 point when it falls below the lambda-z window.
sim_terminal <- dplyr::bind_rows(
  sim_terminal,
  sim_terminal |> dplyr::distinct(id, treatment) |>
    dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

conc_terminal <- PKNCA::PKNCAconc(sim_terminal, Cc ~ time | treatment + id)

# For the terminal window, no doses within the window -- use a synthetic
# "dose at t=0" placeholder so PKNCA has a dose reference for half.life.
dose_terminal <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::group_by(id, treatment) |>
  dplyr::summarise(time = 0, amt = dplyr::last(amt), .groups = "drop")

pk_dose_terminal <- PKNCA::PKNCAdose(dose_terminal, amt ~ time | treatment + id)

intervals_terminal <- data.frame(
  start     = 0,
  end       = 24 * 10,
  half.life = TRUE
)

nca_terminal <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_terminal, pk_dose_terminal, intervals = intervals_terminal)
)
```

```{r nca_compare}
# Assemble a per-treatment simulated summary alongside the paper's
# narrative reference values. The paper gives an accumulation-ratio
# range across dose levels rather than per-dose Cmax / AUC values;
# the derived-half-life value is the single testable claim (85 h,
# Kemal 2026 Section 3.2). Tmax and dose proportionality of exposure
# are qualitative expectations.

sim_day1_summary <- as.data.frame(nca_day1$result) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(sim_median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = sim_median) |>
  dplyr::rename(
    "Treatment"                 = treatment,
    "Simulated Cmax (ng/mL)"    = cmax,
    "Simulated Tmax (h)"        = tmax,
    "Simulated AUC0-24 (ng*h/mL)" = auclast
  )

sim_terminal_summary <- as.data.frame(nca_terminal$result) |>
  dplyr::filter(PPTESTCD == "half.life") |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    "Simulated t1/2 (h, terminal)" = median(PPORRES, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::rename("Treatment" = treatment)

summary_tbl <- dplyr::left_join(sim_day1_summary, sim_terminal_summary, by = "Treatment") |>
  dplyr::mutate(
    "Paper Tmax (h, observed median)" = "~ 2 (Introduction)",
    "Paper t1/2 (h, derived)"         = "85 (Section 3.2)"
  )

knitr::kable(
  summary_tbl,
  caption = "Simulated day-1 Cmax / Tmax / AUC0-24 and terminal-phase half-life per dose arm alongside the paper's reported values. Simulated Tmax should sit near 2 h and simulated t1/2 should be near 85 h; the paper does not tabulate per-dose Cmax / AUC values.",
  digits  = 2
)
```

## Dose proportionality check

The paper states nemtabrutinib bioavailability is "linear at high
concentrations" and the -0.151 low-dose effect on F applies only for
doses < 30 mg (Table 2). All three BELLWAVE-003 doses (45, 65, 80
mg) sit above the 30 mg threshold, so simulated AUC0-24 should scale
proportionally with dose. Ratios of median AUC0-24 relative to the
45 mg arm:

```{r dose-prop}
sim_day1_summary |>
  dplyr::mutate(
    dose_mg = as.numeric(sub(" mg qd", "", `Treatment`)),
    expected_ratio = dose_mg / 45
  ) |>
  dplyr::mutate(
    simulated_ratio = `Simulated AUC0-24 (ng*h/mL)` / `Simulated AUC0-24 (ng*h/mL)`[dose_mg == 45]
  ) |>
  dplyr::select(`Treatment`, dose_mg, simulated_ratio, expected_ratio) |>
  dplyr::rename(
    "Dose (mg)"               = dose_mg,
    "Simulated AUC0-24 ratio" = simulated_ratio,
    "Expected (dose/45)"      = expected_ratio
  ) |>
  knitr::kable(
    caption = "Dose proportionality check across the >= 30 mg linear-F regime.",
    digits  = 3
  )
```

# Exposure-response models

Kemal et al. fit three further models, all logistic regressions run with
`glm` in R rather than in NONMEM, all driven by the same per-subject
exposure metric: `Cavg`, the cumulative on-treatment AUC divided by the
treatment duration, computed by simulating each participant's actual
dosing history through the population PK model above with that
participant's post hoc parameter estimates (Methods 2.3 and 2.4). The
three fits are on different cohorts and different endpoints, and share
no coefficients, so they are packaged as three separate model files that
all point back to this vignette:

| Model | Endpoint | Cohort | Coefficient source |
|-------|----------|--------|--------------------|
| `Kemal_2026_nemtabrutinib_bor` | Best overall response (PR or CR, iwCLL 2018) | CLL/SLL subset, n = 288 | Table S4 (printed) |
| `Kemal_2026_nemtabrutinib_ae` | Any-grade drug-related AE | All treated, n = 578 | Figure 4 left panel (digitized) |
| `Kemal_2026_nemtabrutinib_hypertension` | Any-grade hypertension | All treated, n = 578 | Figure 4 right panel (digitized) |

Time on treatment enters the efficacy model through a saturable term,
`T_TRT / (ET50 + T_TRT)` with ET50 fixed at 200 days, but was screened
and explicitly rejected for both safety endpoints ("Time on therapy was
not found to be a significant covariate for these exposure-safety
relationships", Results 3.5) -- so the safety models carry exposure
only.

## Source trace for the exposure-response models

| Component | Value | Source |
|-----------|-------|--------|
| `logit(p) = b0 + b1*Cavg + b2*[t/(ET50 + t)]` | -- | Table S4 footnote (efficacy model form, verbatim) |
| BOR intercept | -7.3978 | Table S4 (SE 0.9854; Z -7.508; p < 0.001) |
| BOR Cavg slope | 0.0019 per ng/mL | Table S4 (SE 0.0005; Z 3.636; p < 0.001) |
| BOR maximum time effect | 8.7637 | Table S4 (SE 1.1822; Z 7.413; p < 0.001) |
| BOR ET50 (fixed) | 200 days | Table S4 footnote (AIC-optimised, AIC 262.34) |
| Follow-up scenarios drawn | 180 and 360 days | Figure 3 caption |
| AE intercept / slope | 0.4733 / 0.001003 per ng/mL | Figure 4 left panel, **digitized** (see below) |
| Hypertension intercept / slope | -2.3463 / 0.0007645 per ng/mL | Figure 4 right panel, **digitized** (see below) |
| No time term in either safety model | -- | Results 3.5 (prose), confirmed by the logit-linearity of the digitized curves |
| Observed AE rates by exposure quartile | 108/145, 109/145, 107/144, 124/144 | Figure 4 left panel annotations |
| Observed hypertension rates by exposure quartile | 13/145, 22/145, 28/144, 25/144 | Figure 4 right panel annotations |

## Provenance of the two safety models: digitization, and how it was calibrated

The article and its supplement print **no coefficients at all** for the
two exposure-safety models. Table S4 tabulates the exposure-efficacy
model alone; the safety relationships appear only as the fitted curves
of Figure 4. Before falling back to digitization, both Figure 4 panels
were rendered at 600 dpi and inspected for coefficient annotations
printed inside the plot area -- the only numbers there are the observed
per-quartile event fractions listed in the source-trace table above.

Each safety curve was then recovered by rendering its panel at 600 dpi,
calibrating both axes on the tick marks, tracing the black fitted line
at every x position where it could be isolated unambiguously (1042
points for the AE panel, 1091 for hypertension, spanning the full
`Cavg` range of roughly 25 to 2160 ng/mL), and fitting a straight line
to the traced points on the logit scale.

Two properties make this a stronger recovery than reading two points off
a curve.

**The functional form is confirmed rather than assumed.** The traced
points are straight on the logit scale to within 0.010 (AE) and 0.032
(hypertension) on that scale across the whole span -- essentially the
pixel-quantization limit. That is independent evidence both that `Cavg`
enters linearly on the logit scale and that no time-on-treatment term is
hiding in either curve, which is what Results 3.5 asserts in prose.

**The pipeline is calibrated against an answer key inside the same
paper.** Figure 3 draws the exposure-*efficacy* curve at two follow-up
times, and *that* model's coefficients are printed in Table S4. Running
the identical digitization pipeline on Figure 3 therefore has a known
right answer, and the chunk below reproduces that check.

```{r digitization-calibration}
# Curve positions recovered from Figure 3 by the same 600 dpi
# trace-and-fit pipeline used on Figure 4, and the Table S4 values they
# should reproduce.
fig3_digitized <- data.frame(
  follow_up_days     = c(180, 360),
  digitized_intercept = c(-3.2399, -1.7624),
  digitized_slope     = c(0.001922, 0.001924)
)

# Table S4 answer key, propagated through the published model form.
s4_intercept <- -7.3978
s4_slope     <-  0.0019
s4_time_max  <-  8.7637
s4_et50      <-  200

fig3_digitized$key_intercept <-
  s4_intercept + s4_time_max *
  (fig3_digitized$follow_up_days / (s4_et50 + fig3_digitized$follow_up_days))

fig3_digitized |>
  dplyr::mutate(
    intercept_error = digitized_intercept - key_intercept,
    slope_vs_key    = digitized_slope - s4_slope
  ) |>
  dplyr::rename(
    "Follow-up (days)"        = follow_up_days,
    "Digitized intercept"     = digitized_intercept,
    "Digitized slope"         = digitized_slope,
    "Table S4 intercept"      = key_intercept,
    "Intercept error"         = intercept_error,
    "Slope minus Table S4"    = slope_vs_key
  ) |>
  knitr::kable(
    caption = paste(
      "Calibration of the digitization pipeline against the printed",
      "Table S4 coefficients, via the two curves of Figure 3."
    ),
    digits = c(0, 4, 6, 4, 4, 6)
  )

stopifnot(
  # The pipeline recovers each printed intercept to better than 0.01 on
  # the logit scale.
  all(abs(fig3_digitized$digitized_intercept - fig3_digitized$key_intercept) < 0.01),
  # Both curves are drawn from ONE fitted slope, so the two independent
  # traces must agree with each other far more tightly than either
  # agrees with the two-significant-figure printed value.
  abs(diff(fig3_digitized$digitized_slope)) < 1e-5,
  # And they must round to the printed 0.0019.
  all(abs(round(fig3_digitized$digitized_slope, 4) - s4_slope) < 1e-9)
)
```

The recovered slopes, 0.001922 and 0.001924 from two independent curves,
both round to the `0.0019` that Table S4 prints to two significant
figures -- so on this parameter the digitization is marginally *sharper*
than the published table. The two safety-model coefficients therefore
carry an uncertainty of well under 0.01 on the intercept and under 1% on
the slope.

## Replicating Figure 3 (exposure-efficacy)

The efficacy model's coefficients come from Table S4, and Figure 3 is
drawn from the same fit, so evaluating the packaged model against the
digitized Figure 3 curves is a genuine table-versus-figure consistency
check on the paper itself.

```{r er-efficacy}
mod_bor <- rxode2::rxode2(readModelDb("Kemal_2026_nemtabrutinib_bor"))

er_grid <- tidyr::crossing(
  CSS_NEMTA = seq(0, 2000, by = 25),
  T_TRT     = c(180, 360)
) |>
  dplyr::mutate(id = dplyr::row_number(), time = 0)

bor_pred <- rxode2::rxSolve(
  mod_bor,
  events = er_grid,
  keep   = c("CSS_NEMTA", "T_TRT")
) |>
  as.data.frame() |>
  dplyr::mutate(follow_up = factor(
    T_TRT, levels = c(180, 360), labels = c("180 days", "360 days")
  ))

ggplot(bor_pred, aes(CSS_NEMTA, pbor, colour = follow_up)) +
  geom_line(size = 0.9) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_colour_manual(values = c("180 days" = "orange", "360 days" = "brown")) +
  labs(
    x        = "Cavg (ng/mL)",
    y        = "Probability of best overall response",
    colour   = "Simulated treatment time",
    title    = "Exposure-efficacy: BOR vs nemtabrutinib Cavg in CLL/SLL",
    caption  = "Replicates Figure 3 of Kemal 2026 (colours match the published figure)."
  ) +
  theme_bw()
```

```{r er-efficacy-check}
# Independent check: the packaged Table S4 coefficients must reproduce
# the curve positions digitized from the published Figure 3.
fig3_curve_check <- tidyr::crossing(
  CSS_NEMTA = c(0, 500, 1000, 1500, 2000),
  T_TRT     = c(180, 360)
) |>
  dplyr::left_join(fig3_digitized, by = c("T_TRT" = "follow_up_days")) |>
  dplyr::mutate(
    from_figure = plogis(digitized_intercept + digitized_slope * CSS_NEMTA),
    from_model  = plogis(
      s4_intercept + s4_slope * CSS_NEMTA +
        s4_time_max * (T_TRT / (s4_et50 + T_TRT))
    ),
    difference  = from_model - from_figure
  )

fig3_curve_check |>
  dplyr::select(CSS_NEMTA, T_TRT, from_model, from_figure, difference) |>
  dplyr::rename(
    "Cavg (ng/mL)"                = CSS_NEMTA,
    "Treatment time (days)"       = T_TRT,
    "Packaged model (Table S4)"   = from_model,
    "Digitized Figure 3"          = from_figure,
    "Difference"                  = difference
  ) |>
  knitr::kable(
    caption = paste(
      "Packaged exposure-efficacy model against the published Figure 3",
      "curves. Both sides describe the same fit, reached independently",
      "through the supplement's table and through the article's figure."
    ),
    digits = 4
  )

stopifnot(
  # Agreement is limited by Table S4 printing the Cavg slope to only two
  # significant figures: the packaged model uses 0.0019, while the curve
  # in Figure 3 is drawn from the roughly 0.001923 the digitization
  # recovers. Over the full 2000 ng/mL range that rounding alone is worth
  # about 0.046 on the logit scale, so the two sides are required to
  # agree to 0.02 on the probability scale ...
  max(abs(fig3_curve_check$difference)) < 0.02,
  # ... and an order of magnitude better at low exposure, where the
  # rounded slope has little leverage and the comparison is effectively
  # testing the intercept and the time term instead.
  max(abs(fig3_curve_check$difference[fig3_curve_check$CSS_NEMTA <= 500])) < 0.005
)
```

The residual gap is systematic, one-signed, and grows with exposure --
it is the two-significant-figure rounding of the printed slope, not a
transcription error. At zero exposure, where that rounding has no
leverage, the table and the figure agree to 0.0002 on the probability
scale at both follow-up times, which confirms the intercept and the
saturable time term independently.

## Replicating Figure 4 (exposure-safety)

```{r er-safety}
mod_ae  <- rxode2::rxode2(readModelDb("Kemal_2026_nemtabrutinib_ae"))
mod_htn <- rxode2::rxode2(readModelDb("Kemal_2026_nemtabrutinib_hypertension"))

safety_grid <- data.frame(CSS_NEMTA = seq(0, 2160, by = 20)) |>
  dplyr::mutate(id = dplyr::row_number(), time = 0)

ae_pred <- rxode2::rxSolve(mod_ae, events = safety_grid, keep = "CSS_NEMTA") |>
  as.data.frame() |>
  dplyr::transmute(CSS_NEMTA, probability = pae,
                   endpoint = "Any-grade drug-related AE")

htn_pred <- rxode2::rxSolve(mod_htn, events = safety_grid, keep = "CSS_NEMTA") |>
  as.data.frame() |>
  dplyr::transmute(CSS_NEMTA, probability = phtn,
                   endpoint = "Any-grade hypertension")

dplyr::bind_rows(ae_pred, htn_pred) |>
  ggplot(aes(CSS_NEMTA, probability)) +
  geom_line(size = 0.9) +
  facet_wrap(~endpoint) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x       = "Cavg (ng/mL)",
    y       = "Probability of at least one event",
    title   = "Exposure-safety relationships for nemtabrutinib monotherapy",
    caption = "Replicates Figure 4 of Kemal 2026 (n = 578, all indications pooled)."
  ) +
  theme_bw()
```

The sharpest available check on the two safety models does not depend on
digitizing anything further. A logistic regression fit with an intercept
reproduces the overall observed event rate in its own cohort, so the
fitted curve evaluated near the cohort's central exposure must land on
the pooled event rate that Figure 4 prints as per-quartile fractions.
The 65 mg arm supplies 434 of the 578 patients, and its `Cavg`
distribution is centred between roughly 600 and 900 ng/mL, so the curve
is required to bracket the observed rate across that band.

```{r er-safety-check}
safety_obs <- data.frame(
  endpoint = c("Any-grade drug-related AE", "Any-grade hypertension"),
  # Figure 4 panel annotations, summed across the four exposure quartiles.
  events   = c(108 + 109 + 107 + 124, 13 + 22 + 28 + 25),
  n        = c(145 + 145 + 144 + 144, 145 + 145 + 144 + 144)
) |>
  dplyr::mutate(observed_rate = events / n)

safety_band <- dplyr::bind_rows(ae_pred, htn_pred) |>
  dplyr::filter(CSS_NEMTA >= 600, CSS_NEMTA <= 900) |>
  dplyr::group_by(endpoint) |>
  dplyr::summarise(
    predicted_lo = min(probability),
    predicted_hi = max(probability),
    .groups = "drop"
  ) |>
  dplyr::left_join(safety_obs, by = "endpoint")

safety_band |>
  dplyr::select(endpoint, events, n, observed_rate, predicted_lo, predicted_hi) |>
  dplyr::rename(
    "Endpoint"                              = endpoint,
    "Observed events"                       = events,
    "N"                                     = n,
    "Observed rate"                         = observed_rate,
    "Model, Cavg 600 ng/mL"                 = predicted_lo,
    "Model, Cavg 900 ng/mL"                 = predicted_hi
  ) |>
  knitr::kable(
    caption = paste(
      "Digitized exposure-safety models against the pooled observed event",
      "rates printed in Figure 4. The predicted band over the dominant",
      "65 mg arm's central exposure range must contain the observed rate."
    ),
    digits = 4
  )

stopifnot(
  # Each digitized curve reproduces its own cohort's pooled event rate.
  # A mis-recovered intercept would shift the curve off this band
  # immediately: the two bands are only 0.05 and 0.03 wide.
  with(safety_band, all(observed_rate > predicted_lo & observed_rate < predicted_hi)),
  # Exposure raises the risk of both endpoints (the paper's qualitative
  # finding); the AE endpoint is the more common one at every exposure.
  all(diff(ae_pred$probability) > 0),
  all(diff(htn_pred$probability) > 0),
  all(ae_pred$probability > htn_pred$probability)
)
```

Both curves rise across the observed exposure range, reproducing the
paper's finding of a significant trend between `Cavg` and each any-grade
endpoint. Note that Kemal et al. found **no** trend for the Grade 3+
versions of either endpoint, nor for any other safety or tolerability
endpoint tested; those analyses have no published curve and are
therefore not packaged.

# Assumptions and deviations

- Covariate distributions in the virtual cohort were collapsed to the
  reference / typical individual (73.5 kg, 68 y, 41.2 g/L albumin,
  White male with CLL/SLL, no organ impairment, no CYP3A4 modulators,
  no acid-reducing agents) to keep the simulation deterministic across
  the three dose arms. Reproducing the paper's forest-plot covariate
  simulations (Figure 1) would require the full observed covariate
  distributions from BELLWAVE-001 and -003, which are not published.
- The paper does not publish observed concentration-time data or a
  summary NCA table. The comparison chunk above therefore benchmarks
  simulated Tmax (~ 2 h) and terminal half-life (~ 85 h) against the
  paper's narrative values; per-dose Cmax and AUC are shown as
  reference values only.
- Additive residual error (3.19 ng/mL, RSE 101%, 95% CI includes
  zero) was retained in the final model per the paper's full-model
  approach even though its identifiability is very poor.
- IIV variance was computed from the paper's reported CV% via
  omega^2 = log(1 + CV^2), which is the exact log-normal
  transformation matching NONMEM's `$OMEGA` in the supplement code
  (initial estimate 0.164 for CL / 0.0692 for Vc; the final
  parameterization corresponds to the CV% values reported in the
  main-article Table 2).
- The 30 mg dose threshold for the low-dose bioavailability effect is
  encoded as `if (DOSE < 30) dose_lt30 <- 1` using the `DOSE`
  regressor carried on the event table; this matches the NONMEM
  supplement's `IF (DOSE < 30) THEN DOSE_F = 1 + THETA(7)` logic
  (dose-record level, not subject level).

## Exposure-response models

- **Non-paper provenance: both exposure-safety models are digitized.**
  Kemal et al. print no coefficients for the any-grade drug-related AE
  or the any-grade hypertension model anywhere in the article or its
  supplement -- Table S4 covers the exposure-efficacy model only. All
  four values in `Kemal_2026_nemtabrutinib_ae` and
  `Kemal_2026_nemtabrutinib_hypertension` were recovered by tracing the
  fitted curves of Figure 4 at 600 dpi, as described and calibrated in
  the "Provenance of the two safety models" section above. Before
  digitizing, both panels were inspected at 600 dpi for coefficient
  annotations printed inside the plot area; none exist. The
  exposure-efficacy model `Kemal_2026_nemtabrutinib_bor` is **not**
  affected -- every one of its values is printed in Table S4.
- The digitized coefficients carry an uncertainty of better than 0.01 on
  the intercept and better than 1% on the slope, bounded empirically by
  running the same pipeline against Figure 3, whose coefficients are
  printed. They are quoted to four significant figures, which is
  consistent with that bound; they should not be treated as exact to the
  last digit the way the Table S4 values can be.
- No standard errors are carried for the two safety models. The shaded
  bands of Figure 4 are confidence intervals on a fitted probability,
  which do not invert to standard errors on the coefficients without the
  design matrix.
- `Cavg` (`CSS_NEMTA`) enters all three exposure-response models
  **uncentred**, matching the raw `glm` intercepts the paper reports.
  Each intercept is therefore the log-odds at zero exposure, which lies
  far outside the observed data and is not clinically interpretable on
  its own. This differs from the otherwise closely analogous
  `Riggs_2012_albinterferon_svr_gt1`, where the exposure term is centred
  because that source tabulates fitted probabilities rather than
  coefficients.
- `ET50` in the efficacy model is encoded as `fixed(200)` days rather
  than as an estimated parameter, because the Table S4 footnote reports
  it was chosen by minimising AIC over a grid (best AIC 262.34) and then
  held at that value for the analysis. No standard error is reported for
  it.
- Two covariates that Kemal et al. name as expected efficacy predictors,
  prior lines of therapy and TP53 mutation status, are recorded in
  `covariatesDataExcluded` rather than `covariateData`: neither could be
  evaluated in this analysis (all patients had received prior therapy,
  and TP53 status was widely missing), so no coefficient exists. Time on
  treatment is likewise recorded as excluded in both safety models,
  where it was screened and rejected.
- The safety endpoints for which Kemal et al. report **no** exposure
  trend -- Grade 3+ drug-related AEs, Grade 3+ hypertension, any-grade
  and Grade 3+ neutropenia, thrombocytopenia, anemia, infection,
  arrhythmia, diarrhea, rash and hemorrhage, and all tolerability
  endpoints -- have no published curve or coefficient and are not
  packaged.
- **Convention deviations, and why they stand.** Each of the three
  exposure-response models raises the same three `checkModelConventions()`
  warnings, and all three are structural to a logistic endpoint rather
  than defects: `rx.<endpoint>.binom` is auto-created by rxode2 for a
  `dbinom()` endpoint and is reported as a residual-error parameter
  matching neither `propSd`/`addSd` nor a label, and the observation
  variable (`bor`, `ae`, `htn`) is not `Cc` because these models predict
  an event probability, not a drug concentration. Renaming any of them
  would misdescribe the model: there is no residual-error parameter to
  name, because the source fits a fixed-effects `glm` in which all the
  stochasticity lives in the Bernoulli endpoint. The already-shipped
  `Riggs_2012_albinterferon_svr_gt1` -- the direct structural precedent
  for a logistic exposure-response sister model -- raises the identical
  three warnings, so this is established practice in the registry rather
  than a new deviation. `CSS_NEMTA` is registered in
  `inst/references/covariate-columns.md` in this same change and is not
  flagged.
- The exposure-response models consume `CSS_NEMTA` as an input column.
  Generating it end-to-end for a new dosing scenario means simulating
  the PK model above over a full treatment history and forming the
  cumulative AUC divided by treatment duration; the chunks in this
  section instead sweep `CSS_NEMTA` directly across its observed range,
  which is what the published figures plot against.
