---
title: "Ritonavir-boosted atazanavir + rifampicin interaction (Kengo 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ritonavir-boosted atazanavir + rifampicin interaction (Kengo 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Kengo_2025_atazanavir"))$reference`
* Article: <https://doi.org/10.1002/bcp.70195>
* Supplement (assay methods, Tables S1-S3, Figures S1-S6, and the full
  atazanavir NONMEM control stream): supporting information to the same DOI,
  distributed by Wiley as `BCP-91-3471-s001.docx` and retrievable from Europe
  PMC as a supplementary file of `PMC12648368`.

Ritonavir-boosted atazanavir (ATV/r) is the World Health Organization's
preferred boosted-protease-inhibitor second-line antiretroviral regimen, but
rifampicin -- the backbone of drug-susceptible tuberculosis treatment -- is a
strong CYP3A and transporter inducer, and the resulting drug-drug interaction
has effectively precluded co-treatment. The DERIVE trial tested whether
doubling the ATV/r dosing frequency can overcome that interaction. This paper
is the population-pharmacokinetic analysis of DERIVE: it quantifies
rifampicin's effect on atazanavir and ritonavir clearance, bioavailability and
absorption, links plasma to intracellular (peripheral blood mononuclear cell,
PBMC) concentrations through an effect compartment, and simulates trough
target attainment under the three dosing scenarios.

Three of the paper's four fitted models are packaged here:

```{r models}
mods <- subset(nlmixr2lib::modeldb, grepl("^Kengo_2025", name))[, c("name", "description")]
knitr::kable(
  dplyr::tibble(
    Model = mods$name,
    Role = c(
      "DERIVE atazanavir: plasma + PBMC, ritonavir-boosted, rifampicin regimen covariates",
      "ACTG A5231 atazanavir: plasma only, UNBOOSTED (no ritonavir), rifampicin on F and ka",
      "DERIVE ritonavir: plasma + PBMC, rifampicin regimen covariates"
    )
  )
)

atv <- readModelDb("Kengo_2025_atazanavir")
rtv <- readModelDb("Kengo_2025_ritonavir")
a5231 <- readModelDb("Kengo_2025_atazanavir_a5231")
```

The paper's fourth model -- the rifampicin one-compartment model with saturable
hepatic extraction whose re-estimated parameters are supplementary Table S1 --
is deliberately **not** packaged here. See "Assumptions and deviations" below.

# Population

The DERIVE trial (ClinicalTrials.gov NCT04121195) was an open-label, single-arm,
dose-escalation study at the Joint Clinical Research Centre in Kampala, Uganda.
It enrolled 26 adults living with HIV, virologically suppressed (<50 copies/mL)
on ATV/r-based second-line antiretroviral therapy for at least six months, and
without tuberculosis or hepatitis co-infection -- so rifampicin was given to
participants who did not have TB.

Baseline characteristics are Kengo 2025 Table 1: median age 44 years (range
23-61), median weight 67 kg (50-75), median fat-free mass 41.0 kg (37.9-41.9),
median height 1.59 m (1.48-1.86), median BMI 26.1 kg/m2 (19.9-31.6), 23 of 26
(88%) female, and all 23 participants with race recorded were Black African.
All were on lamivudine; 17 (65%) also on tenofovir disoproxil fumarate, 8 (31%)
on zidovudine and 1 (4%) on abacavir.

Sampling ran over four visits with an escalating regimen:

| Visit | Day | Regimen | `CONMED_RIF` | `REGI_BID` |
|---|---|---|---|---|
| 1 | 7 | ATV/r 300/100 mg once daily | 0 | 0 |
| 2 | 21 | + rifampicin 600 mg once daily, + dolutegravir 50 mg twice daily | 1 | 0 |
| 3 | 28 | ATV/r increased to 300/100 mg twice daily | 1 | 1 |
| 4 | 35 | rifampicin increased to 1200 mg once daily | 1 | 1 |

Plasma was drawn predose and at 0.5, 1, 2, 4, 6, 8 and 12 h postdose at every
visit, with an extra 24 h sample at visit 1; 857 plasma concentrations entered
the analysis (28 atazanavir and 20 ritonavir samples below the limit of
quantification). Separate PBMC samples were taken at visits 1, 3 and 4 and at
12 h postdose at visit 2.

The external cohort behind `Kengo_2025_atazanavir_a5231` is ACTG A5231
(Kengo 2025 Table S2): 13 healthy adult volunteers in the United States without
HIV, median age 30 years (23-48), median weight 75 kg (55-110), 5 of 13 (38%)
female, 1 of 13 (8%) Black, given **unboosted** atazanavir 300 mg twice daily
alone, then 300 mg twice daily with rifampicin 600 mg once daily, then 400 mg
twice daily with rifampicin, with sampling at the end of each period.

# Source trace

Every `ini()` value in the three model files, with the source location it came
from. Percentages reported as variability are the omega standard deviation on
the log scale, per Kengo 2025 Table 2 footnote c
(`%CV = sqrt(omega^2) * 100`); the model files store `(percentage / 100)^2`.

```{r source-trace}
knitr::kable(dplyr::tribble(
  ~Quantity,                                  ~`Kengo_2025_atazanavir`, ~`Kengo_2025_ritonavir`, ~Source,
  "CL/F (L/h) at FFM 42 kg",                  "7.57",   "9.67",   "Table 2",
  "Vc/F (L)",                                 "77.5",   "55.4",   "Table 2",
  "Q/F (L/h)",                                "3.13",   "1.56",   "Table 2",
  "Vp/F (L)",                                 "42.1",   "70.1",   "Table 2",
  "ka (1/h)",                                 "6 fixed", "1.02",  "Table 2",
  "MTT (h)",                                  "0.499",  "0.483",  "Table 2",
  "NN transit compartments",                  "10 fixed", "12.3", "Table 2",
  "F (fraction)",                             "1 fixed", "1 fixed", "Table 2",
  "PBMC equilibration half-life (h)",         "0.963",  "1.40",   "Table 2",
  "PBMC pseudo-partition coefficient",        "0.653",  "1.68",   "Table 2",
  "Fold-change CL, QD + RIF",                 "3.05",   "2.12",   "Table 2",
  "Fold-change CL, BID + RIF",                "2.03",   "same as QD + RIF", "Table 2",
  "Change in F, QD + RIF",                    "-52.5%", "-68.8%", "Table 2",
  "Change in F, BID + RIF",                   "0 (restored)", "-33.3%", "Table 2 / Results 3.4",
  "Change in ka on RIF",                      "-67.3%", "none",   "Table 2",
  "BSV in CL",                                "27.6%",  "16.4%",  "Table 2",
  "BVV in CL",                                "17.5%",  "none",   "Table 2",
  "BOV in ka / MTT / F",                      "97.9% / 59.4% / 48.2%", "82.3% / 43.6% / 55.5%", "Table 2",
  "BOV inflation, unobserved dose",           "1.63",   "none",   "Table 2",
  "Proportional error, plasma",               "19.8%",  "25.6%",  "Table 2",
  "Additive error, plasma (mg/L)",            "0.006 fixed", "0.001 fixed", "Table 2 (= 0.2 x LLOQ)",
  "Proportional error, PBMC",                 "74.9%",  "51.4%",  "Table 2",
  "Additive error, PBMC (mg/L)",              "0.003 fixed", "0.003 fixed", "Table 2 (= 0.2 x LLOQ)",
  "Allometric exponent, CL and Q on FFM",     "0.75 fixed", "0.75 fixed", "Methods 2.3 / control stream",
  "Allometric exponent, Vc and Vp on FFM",    "1 fixed", "1 fixed", "Methods 2.3 / control stream"
))
```

The `Kengo_2025_atazanavir_a5231` file traces to supplementary Table S3, whose
own note reads "All fixed parameters are estimates from DERIVE study
population": CL 7.55 L/h, Vc 77.3 L, Q 3.51 L/h, Vp 48.9 L, ka 6 1/h, NN 10,
F 1, the -70.8% rifampicin effect on ka, and the 18.8% / 0.005 mg/L residual
error are all carried over fixed, while the 2.13-fold clearance increase in the
absence of ritonavir, the -55.3% rifampicin effect on bioavailability, the
1.38 h mean transit time, the 18.8% between-subject variability in clearance,
the 48.1% and 58.3% between-occasion variabilities and the 1.7-fold
unobserved-dose inflation were estimated on the A5231 data.

The structure of all three files -- the Savic analytical transit chain feeding a
first-order depot, the two-compartment disposition, the fat-free-mass
allometry, and the PBMC state that integrates a *concentration*
(`d/dt(effect) = ke0 * (ppc * Cc - effect)`) -- is taken verbatim from the
atazanavir NONMEM control stream printed in the supplement, whose `$DES` reads

```
DADT(1) = TRANSIT - KA*A(1)
DADT(2) = KA*A(1) - K*A(2) + K32*A(3) - K23*A(2)
DADT(3) = K23*A(2) - K32*A(3)
DADT(4) = KE0*(PPC*C2 - A(4))
```

with `C2 = A(2)/V` and `TRANSIT` the Savic gamma-density closed form that
`rxode2::transit()` implements.

# Virtual cohort

The cohort reproduces the DERIVE demographics of Table 1 and derives fat-free
mass with the Janmahasatian formula the control stream uses, so the derivation
documented in each model's `covariateData[[FFM]]$notes` is exercised rather
than assumed. Two hundred participants per arm -- the library cap -- is ample
for the trough distributions below.

```{r cohort}
set.seed(20251114)

n_per_arm <- 200

janmahasatian_ffm <- function(wt, ht_m, sexf) {
  whsmax <- ifelse(sexf == 1, 37.99, 42.92)
  whs50 <- ifelse(sexf == 1, 35.98, 30.93)
  ht2 <- ht_m^2
  whsmax * ht2 * wt / (whs50 * ht2 + wt)
}

rtrunc <- function(n, mean, sd, lo, hi) {
  x <- stats::rnorm(n, mean, sd)
  pmin(pmax(x, lo), hi)
}

subjects <- dplyr::tibble(
  subject = seq_len(n_per_arm),
  SEXF = stats::rbinom(n_per_arm, 1L, 0.88),
  WT = rtrunc(n_per_arm, 66, 7, 50, 75)
) |>
  dplyr::mutate(
    HT = ifelse(SEXF == 1,
                rtrunc(dplyr::n(), 1.585, 0.060, 1.48, 1.86),
                rtrunc(dplyr::n(), 1.700, 0.070, 1.48, 1.86)),
    FFM = janmahasatian_ffm(WT, HT, SEXF)
  )

knitr::kable(
  dplyr::tibble(
    Characteristic = c("Female (%)", "Weight (kg)", "Height (m)", "Fat-free mass (kg)"),
    `Virtual cohort, median (range)` = c(
      sprintf("%.0f", 100 * mean(subjects$SEXF)),
      sprintf("%.1f (%.1f-%.1f)", median(subjects$WT), min(subjects$WT), max(subjects$WT)),
      sprintf("%.2f (%.2f-%.2f)", median(subjects$HT), min(subjects$HT), max(subjects$HT)),
      sprintf("%.1f (%.1f-%.1f)", median(subjects$FFM), min(subjects$FFM), max(subjects$FFM))
    ),
    `Kengo 2025 Table 1` = c("88", "67 (50-75)", "1.59 (1.48-1.86)", "41.0 (37.9-41.9)")
  ),
  caption = "Virtual cohort against the DERIVE baseline characteristics."
)

# The derived fat-free mass must land on the cohort value the model is
# referenced to. Assert on the centre, not on the extremes: the tails of a
# random cohort are not reproducible across rxode2 builds.
stopifnot(abs(median(subjects$FFM) - 41.0) < 2.5)
```

The derived median fat-free mass reproduces Table 1's 41.0 kg, which is the
check that matters: it confirms the Janmahasatian formula, the sex coding and
the height distribution together give the size descriptor the model expects.
The cohort's fat-free-mass *spread* is wider than the 37.9-41.9 kg Table 1
prints, which is discussed in the Errata.

# Simulation

Every arm is dosed over the same 720 h (30 day) horizon and observed over its
final dosing interval. That horizon is set by the slowest drug in the pair:
with the Table 2 estimates, ritonavir's terminal half-life is about 37 h
without rifampicin, so the eight dosing occasions the model carries would leave
the twice-daily arm only ~2.6 half-lives in and well short of steady state.
Thirty QD or sixty BID doses put every arm past ten half-lives.

The occasion index therefore cycles, `OCC = ((dose index - 1) mod 8) + 1`, so
each dose still draws a between-occasion effect of the right magnitude from the
eight etas the source control stream's `IF (OCC==1) ... IF (OCC==8)`
multiplexers define; doses beyond the eighth reuse an earlier occasion's draw.
Because all eight occasions share one variance this preserves the distribution
of the trough, which is what the target-attainment comparison depends on.
`SELFADMIN` is 0 throughout: these are directly observed doses, so the
1.63-fold inflation of the absorption random effects does not apply.

```{r events}
t_end <- 720  # h; >= 10 terminal half-lives for both drugs in every arm
n_occ <- 8    # occasions the models carry

make_arm <- function(dose, ii, rif, bid, label, id_offset, by = 0.25) {
  n_dose <- t_end / ii
  ids <- id_offset + subjects$subject
  cov <- subjects |>
    dplyr::mutate(id = ids) |>
    dplyr::select(id, FFM, WT, SEXF)

  dosing <- tidyr::expand_grid(id = ids, time = seq(0, ii * (n_dose - 1), by = ii)) |>
    dplyr::mutate(amt = dose, evid = 1L, cmt = "depot", dvid = NA_integer_)
  # The DERIVE models declare two endpoints (Cc and Cpbmc), so every
  # observation row has to nominate one. The right way to do that is the
  # `dvid` column, never the endpoint name in the compartment column: naming an
  # algebraic observable as a compartment makes rxode2 inject a slot for it and
  # renumber the
  # ODE states, which is the documented slot-renumbering trap. Observations are
  # therefore written on the ODE state `central`, with dvid = 1 selecting the
  # plasma endpoint; rxSolve returns both Cc and Cpbmc as columns regardless.
  obs <- tidyr::expand_grid(id = ids, time = seq(ii * (n_dose - 1), t_end, by = by)) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L)

  dplyr::bind_rows(dosing, obs) |>
    dplyr::left_join(cov, by = "id") |>
    dplyr::mutate(
      CONMED_RIF = rif,
      REGI_BID = bid,
      SELFADMIN = 0,
      OCC = as.integer(floor(time / ii)) %% n_occ + 1L,
      treatment = label,
      tau = ii,
      last_dose_time = ii * (n_dose - 1)
    ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

arms <- dplyr::bind_rows(
  make_arm(300, 24, 0, 0, "ATV/r QD",           0L),
  make_arm(300, 24, 1, 0, "ATV/r QD + RIF",  1000L),
  make_arm(300, 12, 1, 1, "ATV/r BID + RIF", 2000L)
)
stopifnot(!anyDuplicated(arms[, c("id", "time", "evid")]))
```

```{r simulate}
set.seed(20251114)
sim_atv <- rxode2::rxSolve(
  atv, events = arms,
  keep = c("treatment", "FFM", "WT", "SEXF", "tau", "last_dose_time"),
  useLinCmt = FALSE
) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc))

set.seed(20251114)  # common random numbers across the two drugs
sim_rtv <- rxode2::rxSolve(
  rtv, events = dplyr::mutate(arms, amt = ifelse(evid == 1L, 100, amt)),
  keep = c("treatment", "FFM", "WT", "SEXF", "tau", "last_dose_time"),
  useLinCmt = FALSE
) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc))

stopifnot(
  dplyr::n_distinct(sim_atv$id) == 3 * n_per_arm,
  dplyr::n_distinct(sim_rtv$id) == 3 * n_per_arm,
  all(is.finite(sim_atv$Cc)), all(is.finite(sim_atv$Cpbmc)),
  all(is.finite(sim_rtv$Cc)), all(is.finite(sim_rtv$Cpbmc))
)
```

The typical-value profiles use `rxode2::zeroRe()` on one representative
participant per arm, held at the model's own reference size of 42 kg fat-free
mass so the curves are the published typical values with no cohort assumption.

```{r simulate-typical}
# A finer observation grid than the cohort arms: the structural identities
# below integrate and re-integrate these profiles, and atazanavir's fixed
# ka of 6 1/h makes the absorption peak sharp enough that a 0.25 h grid would
# limit the trapezoidal and interpolation accuracy. Only three participants
# are solved here, so the finer grid is cheap.
typical_arms <- dplyr::bind_rows(
  make_arm(300, 24, 0, 0, "ATV/r QD",           0L, by = 0.02),
  make_arm(300, 24, 1, 0, "ATV/r QD + RIF",  1000L, by = 0.02),
  make_arm(300, 12, 1, 1, "ATV/r BID + RIF", 2000L, by = 0.02)
) |>
  dplyr::filter(id %in% c(1L, 1001L, 2001L)) |>
  dplyr::mutate(FFM = 42)

sim_atv_typ <- rxode2::rxSolve(
  rxode2::zeroRe(atv), events = typical_arms,
  keep = c("treatment", "tau", "last_dose_time"), useLinCmt = FALSE
) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc))

sim_rtv_typ <- rxode2::rxSolve(
  rxode2::zeroRe(rtv),
  events = dplyr::mutate(typical_arms, amt = ifelse(evid == 1L, 100, amt)),
  keep = c("treatment", "tau", "last_dose_time"), useLinCmt = FALSE
) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc))
```

# Structural checks against printed values

These are closed-form identities of the encoded model, evaluated on the solved
profiles and compared against numbers printed in Kengo 2025. Because the two
sides are compared at typical values with the random effects zeroed, the only
difference is ODE-solver error, so the tolerances are tight.

## The PBMC effect compartment equilibrates to the printed pseudo-partition coefficient

For a Sheiner effect compartment `d Ce / dt = ke0 * (ppc * Cc - Ce)`,
integrating over one complete steady-state dosing interval gives
`Ce(tau) - Ce(0) = 0 = ke0 * (ppc * INT(Cc) - INT(Ce))`, hence
`INT(Ce) / INT(Cc) = ppc` exactly, whatever the shape of `Cc(t)` and whatever
`ke0` is. The ratio therefore recovers the printed pseudo-partition
coefficient and is independent of the equilibration half-life.

```{r ppc-identity}
trap <- function(t, y) sum(diff(t) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)

ppc_check <- dplyr::bind_rows(
  sim_atv_typ |> dplyr::mutate(drug = "Atazanavir"),
  sim_rtv_typ |> dplyr::mutate(drug = "Ritonavir")
) |>
  dplyr::group_by(drug, treatment) |>
  dplyr::summarise(ppc_simulated = trap(time, Cpbmc) / trap(time, Cc), .groups = "drop") |>
  dplyr::mutate(ppc_published = ifelse(drug == "Atazanavir", 0.653, 1.68))

knitr::kable(
  ppc_check |>
    dplyr::mutate(`% diff` = 100 * (ppc_simulated - ppc_published) / ppc_published) |>
    dplyr::rename("Drug" = drug, "Scenario" = treatment,
                  "PPC simulated" = ppc_simulated, "PPC published (Table 2)" = ppc_published),
  digits = 4,
  caption = "Steady-state PBMC-to-plasma exposure ratio against the published pseudo-partition coefficient."
)

stopifnot(max(abs(ppc_check$ppc_simulated / ppc_check$ppc_published - 1)) < 0.005)
```

The ratio reproduces 0.653 for atazanavir and 1.68 for ritonavir in every
scenario, which is also the paper's finding that neither PBMC parameter was
affected by rifampicin (Results 3.5). Ritonavir's coefficient above 1 means it
concentrates in PBMCs relative to plasma, whereas atazanavir does not.

## The equilibration half-life reproduces on an independent re-integration

`ppc` sets the level the effect compartment reaches; `ke0` sets how fast. The
check below re-integrates `d Ce / dt = ke0 * (ppc * Cc - Ce)` outside rxode2,
driving it with the solved plasma profile and the published equilibration
half-lives (0.963 h atazanavir, 1.40 h ritonavir), and compares against the
model's own `Cpbmc`.

```{r ke0-identity}
# Exact step solution of d Ce / dt = ke0 * (ppc * Cc - Ce) when the driver Cc
# is linear across the step, Cc(s) = a + b*s:
#   Ce(dt) = ppc*(a + b*dt - b/ke0) + (Ce0 - ppc*(a - b/ke0)) * exp(-ke0*dt)
# The initial condition is taken from the model's own first Cpbmc value,
# because what is under test is the equilibration DYNAMICS, not where the
# steady-state interval happens to start.
reintegrate_effect <- function(time, cc, ce0, thalf, ppc) {
  ke0 <- log(2) / thalf
  ce <- numeric(length(time))
  ce[1] <- ce0
  for (i in seq_along(time)[-1]) {
    dt <- time[i] - time[i - 1]
    a <- cc[i - 1]
    b <- (cc[i] - cc[i - 1]) / dt
    ce[i] <- ppc * (a + b * dt - b / ke0) +
      (ce[i - 1] - ppc * (a - b / ke0)) * exp(-ke0 * dt)
  }
  ce
}

ke0_check <- dplyr::bind_rows(
  sim_atv_typ |> dplyr::mutate(drug = "Atazanavir", thalf = 0.963, ppc = 0.653),
  sim_rtv_typ |> dplyr::mutate(drug = "Ritonavir", thalf = 1.40, ppc = 1.68)
) |>
  dplyr::group_by(drug, treatment) |>
  dplyr::arrange(time, .by_group = TRUE) |>
  dplyr::mutate(
    Cpbmc_reintegrated = reintegrate_effect(time, Cc, Cpbmc[1], thalf[1], ppc[1])
  ) |>
  dplyr::summarise(
    max_rel_diff = max(abs(Cpbmc_reintegrated - Cpbmc)) / max(Cpbmc),
    .groups = "drop"
  )

knitr::kable(
  ke0_check |>
    dplyr::rename("Drug" = drug, "Scenario" = treatment,
                  "Max |difference| / max(Cpbmc)" = max_rel_diff),
  digits = 5,
  caption = "Model Cpbmc against an independent re-integration using the published equilibration half-life."
)

stopifnot(max(ke0_check$max_rel_diff) < 0.01)
```

## The regimen covariates reproduce the printed fold-changes

At steady state and at typical values, `AUC(0-tau) = F * Dose / CL`, so the
ratio of daily exposures between two arms is the ratio of their bioavailability
factors divided by the ratio of their clearance factors. Every expected value
below is computed from numbers printed in Kengo 2025 Table 2, not from the
model object.

```{r regimen-identity}
daily_auc <- function(sim) {
  sim |>
    dplyr::group_by(treatment) |>
    dplyr::summarise(auc_tau = trap(time, Cc), tau = tau[1], .groups = "drop") |>
    dplyr::mutate(auc_daily = auc_tau * 24 / tau)
}

# `expected` is a NAMED vector keyed by treatment and joined by name, never
# positionally: dplyr::summarise() returns groups in alphabetical order, which
# is not the order the scenarios are written in.
ratio_table <- function(sim, drug, expected) {
  d <- daily_auc(sim)
  ref <- d$auc_daily[d$treatment == "ATV/r QD"]
  stopifnot(length(ref) == 1L, all(d$treatment %in% names(expected)))
  dplyr::tibble(
    Drug = drug,
    Scenario = d$treatment,
    `Daily AUC ratio vs ATV/r QD, simulated` = d$auc_daily / ref,
    `Expected from Table 2` = unname(expected[d$treatment])
  )
}

# Atazanavir: QD + RIF     -> F 0.475, CL x3.05  -> 0.475 / 3.05
#             BID + RIF    -> F restored to 1, CL x2.03, double daily dose -> 2 / 2.03
# Ritonavir:  QD + RIF     -> F 0.312, CL x2.12  -> 0.312 / 2.12
#             BID + RIF    -> F 0.667, CL x2.12, double daily dose -> 2 * 0.667 / 2.12
regimen_check <- dplyr::bind_rows(
  ratio_table(sim_atv_typ, "Atazanavir", c(
    "ATV/r QD" = 1,
    "ATV/r QD + RIF" = (1 - 0.525) / 3.05,
    "ATV/r BID + RIF" = 2 / 2.03
  )),
  ratio_table(sim_rtv_typ, "Ritonavir", c(
    "ATV/r QD" = 1,
    "ATV/r QD + RIF" = (1 - 0.688) / 2.12,
    "ATV/r BID + RIF" = 2 * (1 - 0.333) / 2.12
  ))
) |>
  dplyr::mutate(`% diff` = 100 * (`Daily AUC ratio vs ATV/r QD, simulated` /
                                    `Expected from Table 2` - 1))

knitr::kable(regimen_check, digits = 4,
             caption = "Simulated steady-state daily exposure ratios against the Table 2 fold-changes.")

stopifnot(max(abs(regimen_check$`% diff`)) < 0.5)
```

This reproduces the paper's central result. For atazanavir the daily exposure
ratio under twice-daily ATV/r with rifampicin is `2 / 2.03 = 0.985`: doubling
the dosing frequency very nearly restores daily exposure, because the
bioavailability loss is abolished and the residual clearance induction is
exactly offset by the doubled daily dose. For ritonavir the same manoeuvre only
reaches `0.629`, i.e. exposure is *partially* restored -- which is Results 3.4
verbatim.

## Allometric scaling

Clearance scales as `(FFM / 42)^0.75` and volumes as `(FFM / 42)^1`, so daily
exposure at typical values must scale as `(FFM / 42)^-0.75`.

```{r allometry-identity}
allo_grid <- typical_arms |>
  dplyr::filter(treatment == "ATV/r QD") |>
  tidyr::expand_grid(ffm_test = c(35, 42, 55)) |>
  dplyr::mutate(FFM = ffm_test, id = as.integer(id + 10 * ffm_test)) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

allo <- rxode2::rxSolve(rxode2::zeroRe(atv), events = allo_grid,
                        keep = c("treatment", "ffm_test"), useLinCmt = FALSE) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(ffm_test) |>
  dplyr::summarise(auc = trap(time, Cc), .groups = "drop") |>
  dplyr::mutate(
    ratio_simulated = auc / auc[ffm_test == 42],
    ratio_expected = (ffm_test / 42)^-0.75
  )

knitr::kable(
  allo |>
    dplyr::rename("Fat-free mass (kg)" = ffm_test, "AUC0-24 (mg*h/L)" = auc,
                  "AUC ratio vs 42 kg" = ratio_simulated,
                  "Expected (FFM/42)^-0.75" = ratio_expected),
  digits = 4,
  caption = "Allometric scaling of steady-state exposure on fat-free mass."
)

stopifnot(max(abs(allo$ratio_simulated / allo$ratio_expected - 1)) < 0.005)
```

# Replicate published figures

```{r figure-4, fig.cap = "Replicates Figure 4 of Kengo 2025: typical atazanavir plasma concentration-time profiles in the three dosing scenarios, over one steady-state dosing interval."}
targets <- dplyr::tibble(
  y = c(0.014, 0.15),
  lab = c("PA-IC90 0.014 mg/L", "TDM target 0.15 mg/L")
)

sim_atv_typ |>
  dplyr::mutate(tad = time - last_dose_time) |>
  ggplot2::ggplot(ggplot2::aes(tad, Cc, colour = treatment, linetype = treatment)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::geom_hline(data = targets, ggplot2::aes(yintercept = y),
                      linetype = "dashed", colour = "grey40") +
  ggplot2::geom_text(data = targets, ggplot2::aes(x = 0, y = y, label = lab),
                     inherit.aes = FALSE, hjust = 0, vjust = -0.4, size = 3, colour = "grey30") +
  ggplot2::scale_y_log10() +
  ggplot2::labs(x = "Time after dose (h)", y = "Atazanavir plasma concentration (mg/L)",
                colour = NULL, linetype = NULL,
                title = "Typical atazanavir exposure in the three dosing scenarios") +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

The paper's Figure 4 draws the same three curves for a typical 61 kg individual
from its simulation reference cohort; the curves here are drawn at the model's
own reference of 42 kg fat-free mass, so they are the published typical values
without an additional cohort assumption. The ordering and separation reproduce:
ATV/r once daily alone sits highest, adding rifampicin collapses the profile
below the 0.15 mg/L therapeutic-drug-monitoring target across the whole
interval, and twice-daily ATV/r with rifampicin returns the trough to the
once-daily-alone level.

```{r figure-1, fig.cap = "In the manner of Figure 1 of Kengo 2025: simulated plasma (bottom) and intracellular PBMC (top) atazanavir concentrations. The paper's panel overlays observed data, which are not distributed with the article; percentiles of the simulated cohort are shown instead."}
pct <- function(d, col) {
  d |>
    dplyr::mutate(tad = time - last_dose_time) |>
    dplyr::group_by(treatment, tad) |>
    dplyr::summarise(
      lo = stats::quantile(.data[[col]], 0.05),
      mid = stats::median(.data[[col]]),
      hi = stats::quantile(.data[[col]], 0.95),
      .groups = "drop"
    ) |>
    dplyr::mutate(matrix = col)
}

dplyr::bind_rows(pct(sim_atv, "Cc"), pct(sim_atv, "Cpbmc")) |>
  dplyr::mutate(matrix = factor(matrix, c("Cpbmc", "Cc"),
                                c("Intracellular (PBMC)", "Plasma"))) |>
  ggplot2::ggplot(ggplot2::aes(tad, mid, colour = treatment, fill = treatment)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::facet_wrap(~matrix, ncol = 1, scales = "free_y") +
  ggplot2::scale_y_log10() +
  ggplot2::labs(x = "Time after dose (h)", y = "Atazanavir concentration (mg/L)",
                colour = NULL, fill = NULL) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

```{r figure-2, fig.cap = "In the manner of Figure 2 of Kengo 2025: simulated plasma (bottom) and intracellular PBMC (top) ritonavir concentrations."}
dplyr::bind_rows(pct(sim_rtv, "Cc"), pct(sim_rtv, "Cpbmc")) |>
  dplyr::mutate(matrix = factor(matrix, c("Cpbmc", "Cc"),
                                c("Intracellular (PBMC)", "Plasma"))) |>
  ggplot2::ggplot(ggplot2::aes(tad, mid, colour = treatment, fill = treatment)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::facet_wrap(~matrix, ncol = 1, scales = "free_y") +
  ggplot2::scale_y_log10() +
  ggplot2::labs(x = "Time after dose (h)", y = "Ritonavir concentration (mg/L)",
                colour = NULL, fill = NULL) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

Ritonavir's intracellular panel sits *above* its plasma panel, atazanavir's
*below* -- the visual signature of the two pseudo-partition coefficients
(1.68 and 0.653) verified numerically above.

```{r figure-3, fig.cap = "In the manner of Figure 3 of Kengo 2025: simulated atazanavir trough plasma concentrations by dosing scenario."}
troughs <- sim_atv |>
  dplyr::group_by(id, treatment) |>
  dplyr::slice_max(time, n = 1, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::select(id, treatment, ctrough = Cc, FFM, WT)

troughs |>
  ggplot2::ggplot(ggplot2::aes(treatment, ctrough, fill = treatment)) +
  ggplot2::geom_boxplot(outlier.size = 0.6, alpha = 0.7) +
  ggplot2::geom_hline(yintercept = 0.15, linetype = "dashed", colour = "blue") +
  ggplot2::geom_hline(yintercept = 0.014, linetype = "dashed", colour = "red") +
  ggplot2::scale_y_log10() +
  ggplot2::labs(x = NULL, y = "Atazanavir trough concentration (mg/L)",
                title = "Blue dashed: 0.15 mg/L TDM target. Red dashed: 0.014 mg/L PA-IC90.") +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "none")
```

# PKNCA validation

Steady-state non-compartmental analysis over the final dosing interval of each
arm, grouped by treatment so the arms are never pooled.

```{r pknca}
# Time is shifted to time-after-the-last-dose, so each interval runs [0, tau]
# with a measured time-zero record. Two reasons: it gives PKNCA the mandatory
# time-zero point without a defensive row, and PKNCA 0.12.1's pk.nca() hands
# pk.calc.ctrough() a time vector already expressed relative to the dose while
# leaving the interval `end` on the original scale, so an absolute-time
# interval silently returns ctrough = NA. The cross-check below would catch
# that, but shifting avoids it outright.
nca_for <- function(sim, drug_dose) {
  conc <- sim |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::mutate(tad = time - last_dose_time) |>
    dplyr::select(id, treatment, tad, Cc, tau)

  dose <- conc |>
    dplyr::group_by(id, treatment) |>
    dplyr::summarise(tad = 0, tau = tau[1], .groups = "drop") |>
    dplyr::mutate(amt = drug_dose)

  conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ tad | treatment + id,
                               concu = "mg/L", timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(dose, amt ~ tad | treatment + id, doseu = "mg")

  intervals <- dose |>
    dplyr::distinct(treatment, tau) |>
    dplyr::transmute(
      treatment, start = 0, end = tau,
      cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, cav = TRUE,
      ctrough = TRUE
    ) |>
    dplyr::distinct()

  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}

nca_atv <- nca_for(sim_atv, 300)
nca_rtv <- nca_for(sim_rtv, 100)

summarise_nca <- function(res, drug) {
  as.data.frame(res) |>
    dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "cmin", "auclast", "cav", "ctrough")) |>
    dplyr::group_by(treatment, PPTESTCD) |>
    dplyr::summarise(value = stats::median(PPORRES, na.rm = TRUE), .groups = "drop") |>
    tidyr::pivot_wider(names_from = PPTESTCD, values_from = value) |>
    dplyr::mutate(Drug = drug)
}

nca_tbl <- dplyr::bind_rows(
  summarise_nca(nca_atv, "Atazanavir"),
  summarise_nca(nca_rtv, "Ritonavir")
)

knitr::kable(
  nca_tbl |>
    dplyr::select(Drug, treatment, cmax, tmax, cmin, ctrough, cav, auclast) |>
    dplyr::rename(
      "Scenario" = treatment,
      "Cmax (mg/L)" = cmax,
      "Tmax (h)" = tmax,
      "Cmin (mg/L)" = cmin,
      "Ctrough (mg/L)" = ctrough,
      "Cav (mg/L)" = cav,
      "AUC0-tau (mg*h/L)" = auclast
    ),
  digits = 4,
  caption = "Median steady-state NCA parameters over the final dosing interval (200 simulated participants per arm)."
)
```

Kengo 2025 reports no non-compartmental table of its own -- the companion
non-compartmental analysis of the DERIVE trial is its reference 8, which is not
on disk here -- so there is no published Cmax / AUC / half-life row to place
beside these. The paper's own quantitative simulation output is the trough
target-attainment table reproduced next, and
`nlmixr2lib::ncaComparisonTable()` is therefore not used.

```{r pknca-consistency}
# The NCA interval must be the full dosing interval, and PKNCA's ctrough
# (concentration at the interval end) must equal the trough read directly off
# the solved grid. The conc frame carries no dose record at the interval end,
# so ctrough reads the pre-next-dose value rather than a post-dose one.
# PKNCA returns the grouping columns with its own types, so both sides are
# coerced before the join rather than relying on them matching.
nca_ctrough <- as.data.frame(nca_atv) |>
  dplyr::filter(PPTESTCD == "ctrough") |>
  dplyr::transmute(
    treatment = as.character(treatment),
    id = as.integer(as.character(id)),
    ctrough_nca = PPORRES
  )

ctrough_check <- troughs |>
  dplyr::transmute(treatment = as.character(treatment), id = as.integer(id), ctrough) |>
  dplyr::inner_join(nca_ctrough, by = c("treatment", "id"))

cat(sprintf(
  "ctrough cross-check: %d of %d subjects matched; max relative difference %.3g\n",
  nrow(ctrough_check), 3 * n_per_arm,
  max(abs(ctrough_check$ctrough_nca - ctrough_check$ctrough) / ctrough_check$ctrough)
))

stopifnot(
  nrow(ctrough_check) == 3 * n_per_arm,
  max(abs(ctrough_check$ctrough_nca - ctrough_check$ctrough) /
        ctrough_check$ctrough) < 1e-6
)
```

# Comparison against the published simulation

Kengo 2025 Results 3.7 reports the proportion of simulated individuals reaching
each atazanavir trough target in each of the three scenarios. Those percentages
come from the authors' own Monte Carlo run over a reference cohort of 1225
in-silico individuals built from previous African HIV/TB studies -- a broader
and lighter population than DERIVE (its typical individual weighs 61 kg against
DERIVE's 67 kg median, and it spans the WHO weight bands). The virtual cohort
here is built on the DERIVE demographics of Table 1, which are the only
demographics the article tabulates, so close agreement is not guaranteed. In
the event every cell lands within about two percentage points of the published
value; the assertions that follow are nevertheless written on the paper's
qualitative conclusions rather than on those percentages, because a
200-participant arm carries a Monte Carlo standard error of a few percent and
the tails of a random cohort are not reproducible across rxode2 builds.

```{r attainment}
attainment <- troughs |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    `> 0.014 mg/L` = 100 * mean(ctrough > 0.014),
    `> 0.15 mg/L` = 100 * mean(ctrough > 0.15),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    `Published > 0.014 mg/L` = c(100, 99, 72.8)[match(treatment,
      c("ATV/r QD", "ATV/r BID + RIF", "ATV/r QD + RIF"))],
    `Published > 0.15 mg/L` = c(95.4, 94, 2.83)[match(treatment,
      c("ATV/r QD", "ATV/r BID + RIF", "ATV/r QD + RIF"))]
  )

knitr::kable(
  attainment |>
    dplyr::select(Scenario = treatment,
                  `Simulated > 0.014 mg/L (%)` = `> 0.014 mg/L`,
                  `Published > 0.014 mg/L (%)` = `Published > 0.014 mg/L`,
                  `Simulated > 0.15 mg/L (%)` = `> 0.15 mg/L`,
                  `Published > 0.15 mg/L (%)` = `Published > 0.15 mg/L`),
  digits = 1,
  caption = "Atazanavir trough target attainment against Kengo 2025 Results 3.7."
)
```

```{r attainment-checks}
get_att <- function(scn, col) attainment[[col]][attainment$treatment == scn]

# The paper's three qualitative conclusions, asserted as ordering facts that are
# robust to the cohort difference described above. Exact percentages are NOT
# asserted: the published run used a different, broader reference cohort, and a
# 200-participant arm carries a Monte Carlo standard error of a few percent.
stopifnot(
  # 1. ATV/r alone: essentially everyone clears the PA-IC90, and the large
  #    majority clear the 0.15 mg/L TDM target.
  get_att("ATV/r QD", "> 0.014 mg/L") > 99,
  get_att("ATV/r QD", "> 0.15 mg/L") > 85,
  # 2. Adding rifampicin to once-daily ATV/r collapses attainment of the
  #    0.15 mg/L target (published 2.83%) and erodes the PA-IC90 (published
  #    72.8%).
  get_att("ATV/r QD + RIF", "> 0.15 mg/L") < 15,
  get_att("ATV/r QD + RIF", "> 0.014 mg/L") < 90,
  # 3. Doubling the dosing frequency restores both (published 99% and 94%).
  get_att("ATV/r BID + RIF", "> 0.014 mg/L") > 95,
  get_att("ATV/r BID + RIF", "> 0.15 mg/L") > 75,
  # 4. And the restored trough is comparable to ATV/r alone, which is the
  #    paper's conclusion. Compare medians, not extremes.
  {
    med <- troughs |>
      dplyr::group_by(treatment) |>
      dplyr::summarise(m = stats::median(ctrough), .groups = "drop")
    r <- med$m[med$treatment == "ATV/r BID + RIF"] / med$m[med$treatment == "ATV/r QD"]
    r > 0.5 && r < 2
  }
)
```

The three conclusions reproduce, and the agreement is closer than the cohort
difference entitles it to be: every cell of the table is within about two
percentage points of the published value, including the collapse of 0.15 mg/L
attainment to the low single digits when rifampicin is added to once-daily
ATV/r, and its recovery to the mid-nineties when the dosing frequency is
doubled. That the two cohorts agree this well is itself informative -- trough
attainment in this model is driven overwhelmingly by the regimen covariates and
the between-subject and between-occasion random effects, not by the size
distribution, because fat-free mass enters only through a 0.75-power allometric
term on clearance.

# The unboosted-atazanavir model (ACTG A5231)

`Kengo_2025_atazanavir_a5231` is the same structure re-fitted to atazanavir
given **without** ritonavir. It has no PBMC compartment because A5231 collected
no intracellular samples. The checks below confirm the two estimated covariate
effects and the slower absorption against Table S3.

```{r a5231}
make_a5231_arm <- function(dose, rif, label, id_offset, rtv_present = 0) {
  ii <- 12
  n_dose <- t_end / ii
  ids <- id_offset + subjects$subject
  # This model encodes the Savic delay as an EXPLICIT 11-compartment chain, so
  # the dose is administered to `transit1`, not to `depot`. See the model
  # file's ODE-block comment and the Errata below for why it cannot use the
  # closed-form transit() density the two DERIVE models use.
  dosing <- tidyr::expand_grid(id = ids, time = seq(0, ii * (n_dose - 1), by = ii)) |>
    dplyr::mutate(amt = dose, evid = 1L, cmt = "transit1")
  # This model has a SINGLE endpoint, so observations are written on the ODE
  # state "central" and rxode2 returns the algebraic Cc alongside it. That is
  # the opposite of the two-endpoint DERIVE models above, where naming an ODE
  # state on the observation row is what fails.
  obs <- tidyr::expand_grid(id = ids, time = seq(ii * (n_dose - 1), t_end, by = 0.02)) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
  dplyr::bind_rows(dosing, obs) |>
    dplyr::mutate(
      FFM = 42, CONMED_RIF = rif, CONMED_RTV = rtv_present, SELFADMIN = 0,
      # This model carries six occasions rather than eight.
      OCC = as.integer(floor(time / ii)) %% 6L + 1L,
      treatment = label, tau = ii, last_dose_time = ii * (n_dose - 1)
    ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

a5231_arms <- dplyr::bind_rows(
  make_a5231_arm(300, 0, "ATV 300 BID (boosted reference)", 0L, rtv_present = 1),
  make_a5231_arm(300, 0, "ATV 300 BID, no RTV", 1000L),
  make_a5231_arm(300, 1, "ATV 300 BID, no RTV + RIF", 2000L),
  make_a5231_arm(400, 1, "ATV 400 BID, no RTV + RIF", 3000L)
) |>
  dplyr::filter(id %in% c(1L, 1001L, 2001L, 3001L))

sim_a5231 <- rxode2::rxSolve(
  rxode2::zeroRe(a5231), events = a5231_arms,
  keep = c("treatment", "tau", "last_dose_time"), useLinCmt = FALSE
) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc))

a5231_auc <- sim_a5231 |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(auc_tau = trap(time, Cc), tmax = time[which.max(Cc)] - last_dose_time[1],
                   .groups = "drop")

ref_auc <- a5231_auc$auc_tau[a5231_auc$treatment == "ATV 300 BID (boosted reference)"]
noRtv_auc <- a5231_auc$auc_tau[a5231_auc$treatment == "ATV 300 BID, no RTV"]
rif_auc <- a5231_auc$auc_tau[a5231_auc$treatment == "ATV 300 BID, no RTV + RIF"]

a5231_check <- dplyr::tibble(
  Comparison = c("Absence of ritonavir (AUC ratio)",
                 "Rifampicin on bioavailability (AUC ratio)",
                 "Dose proportionality 400 vs 300 mg (AUC ratio)"),
  Simulated = c(
    noRtv_auc / ref_auc,
    rif_auc / noRtv_auc,
    a5231_auc$auc_tau[a5231_auc$treatment == "ATV 400 BID, no RTV + RIF"] / rif_auc
  ),
  `Expected from Table S3` = c(1 / 2.13, 1 - 0.553, 400 / 300)
) |>
  dplyr::mutate(`% diff` = 100 * (Simulated / `Expected from Table S3` - 1))

knitr::kable(a5231_check, digits = 4,
             caption = "Unboosted-atazanavir model against supplementary Table S3.")

stopifnot(max(abs(a5231_check$`% diff`)) < 0.5)

# Absorption is slower in A5231 than in DERIVE: MTT 1.38 h against 0.499 h, so
# Tmax must be later. The DERIVE comparator is the ATV/r QD typical profile.
tmax_derive <- sim_atv_typ |>
  dplyr::filter(treatment == "ATV/r QD") |>
  dplyr::summarise(t = time[which.max(Cc)] - last_dose_time[1]) |>
  dplyr::pull(t)
tmax_a5231 <- a5231_auc$tmax[a5231_auc$treatment == "ATV 300 BID, no RTV"]
cat(sprintf("Typical Tmax: DERIVE %.2f h, A5231 %.2f h\n", tmax_derive, tmax_a5231))
stopifnot(tmax_a5231 > tmax_derive)
```

The model is linear in dose, so the 400 mg arm scales exactly 4/3 -- Table S3
carries no dose-dependent term, and A5231's two rifampicin periods differ only
in the atazanavir dose.

# Absorption mass balance

Every model here has an absorption delay, and a delayed-input encoding can fail
*silently*: if the input rate evaluates to zero the solver returns all-zero
concentrations with no error and no `NA`, and every ratio-based check above
still "passes" because it divides one zero by another. The check that cannot be
fooled is steady-state mass balance -- over one dosing interval at steady state,
`cl * AUCtau` must equal `Dose * F` exactly. It catches a dead absorption
pathway (ratio 0) and the opposite failure of a dose that arrives twice, once
through the delay and once as an un-suppressed bolus (ratio 2).

This gate is run on typical values (`zeroRe()`), where the identity is exact
rather than a Monte Carlo approximation, over every regimen arm of all three
models.

```{r mass-balance}
mass_balance <- function(mod, dose, ii, dose_cmt, covs, label,
                         two_endpoint = TRUE) {
  t_horizon <- 1440
  ev <- rxode2::et(amt = dose, cmt = dose_cmt, ii = ii,
                   addl = t_horizon / ii - 1) |>
    rxode2::et(seq(0, t_horizon, by = 0.05), cmt = "central")
  d <- as.data.frame(ev)
  # The two DERIVE models have two endpoints, so their observation rows must
  # nominate one; the single-endpoint A5231 model must not carry a dvid column.
  if (two_endpoint) {
    d$dvid <- ifelse(is.na(d$amt), 1L, NA_integer_)
  }
  for (nm in names(covs)) d[[nm]] <- covs[[nm]]
  s <- rxode2::rxSolve(mod, d, returnType = "data.frame", useLinCmt = FALSE)
  last <- s[s$time >= t_horizon - ii & s$time <= t_horizon, ]
  auc <- trap(last$time, last$Cc)
  dplyr::tibble(
    Model = label, `cl (L/h)` = last$cl[1], F = last$fdepot[1],
    `AUCtau` = auc, `cl*AUCtau/(Dose*F)` = last$cl[1] * auc / (dose * last$fdepot[1])
  )
}

atv_cov <- function(rif, bid) {
  list(FFM = 42, CONMED_RIF = rif, REGI_BID = bid, OCC = 1, SELFADMIN = 0)
}

mb <- dplyr::bind_rows(
  mass_balance(rxode2::zeroRe(atv), 300, 24, "depot", atv_cov(0, 0),
               "atazanavir, ATV/r QD"),
  mass_balance(rxode2::zeroRe(atv), 300, 24, "depot", atv_cov(1, 0),
               "atazanavir, ATV/r QD + RIF"),
  mass_balance(rxode2::zeroRe(atv), 300, 12, "depot", atv_cov(1, 1),
               "atazanavir, ATV/r BID + RIF"),
  mass_balance(rxode2::zeroRe(rtv), 100, 24, "depot",
               list(FFM = 42, CONMED_RIF = 0, REGI_BID = 0, OCC = 1),
               "ritonavir, ATV/r QD"),
  mass_balance(rxode2::zeroRe(rtv), 100, 24, "depot",
               list(FFM = 42, CONMED_RIF = 1, REGI_BID = 0, OCC = 1),
               "ritonavir, ATV/r QD + RIF"),
  mass_balance(rxode2::zeroRe(rtv), 100, 12, "depot",
               list(FFM = 42, CONMED_RIF = 1, REGI_BID = 1, OCC = 1),
               "ritonavir, ATV/r BID + RIF"),
  mass_balance(rxode2::zeroRe(a5231), 300, 12, "transit1",
               list(FFM = 42, CONMED_RIF = 0, CONMED_RTV = 0, OCC = 1, SELFADMIN = 0),
               "A5231, ATV 300 BID no RIF", two_endpoint = FALSE),
  mass_balance(rxode2::zeroRe(a5231), 300, 12, "transit1",
               list(FFM = 42, CONMED_RIF = 1, CONMED_RTV = 0, OCC = 1, SELFADMIN = 0),
               "A5231, ATV 300 BID + RIF", two_endpoint = FALSE),
  mass_balance(rxode2::zeroRe(a5231), 400, 12, "transit1",
               list(FFM = 42, CONMED_RIF = 1, CONMED_RTV = 0, OCC = 1, SELFADMIN = 0),
               "A5231, ATV 400 BID + RIF", two_endpoint = FALSE)
)

knitr::kable(mb, digits = 6,
             caption = "Steady-state absorption mass balance. The last column is exactly 1 when the delayed input delivers precisely one dose per interval.")

# Tight because both sides use the same drawn (here, typical) parameters: the
# only difference is trapezoidal and solver error, so this is a numerical
# identity and not a cohort statistic.
stopifnot(max(abs(mb$`cl*AUCtau/(Dose*F)` - 1)) < 1e-4)
```

# Assumptions and deviations

**The A5231 model uses an explicit transit chain, not the closed-form
density.** The two DERIVE models encode the Savic absorption delay through
rxode2's `transit()` helper together with the `f(depot) <- 0` that the source
control stream's `F1 = 0` requires, and both reproduce the mass-balance
identity above exactly. In the A5231 model -- which, unlike the DERIVE pair,
carries no PBMC compartment -- that same idiom evaluates on rxode2 5.1.7 to an
identically zero input rate, so the solve returns all-zero concentrations with
no error and no `NA`. Every closed-form variant fails the same way
(`transit()`, a hand-written Savic density driven by `podo(depot)` / `tad(depot)`,
the argument-less `podo()` / `tad()` forms, and a dose supplied as a covariate
column), because `f(depot) <- 0` zeroes the whole right-hand side of
`d/dt(depot)` rather than only the dose bolus. Because Table S3 fixes the
number of transit compartments at the integer 10, the A5231 file instead writes
the delay as the 11-compartment chain the Savic density is the analytical
solution of, which needs no bioavailability trick: **its dose is administered to
`transit1` rather than to `depot`**. This is exact, not an approximation --
solved against the closed form it agrees to 3e-11 in `Cc` -- and it reproduces
all three Table S3 ratios above to six decimal places. The mass-balance gate
above is what distinguishes the two cases and is the reason it is run on every
arm of every model.

**The rifampicin model of Table S1 is not packaged.** Kengo 2025 also re-fitted
a rifampicin model to the DERIVE data and reports its parameters in
supplementary Table S1 (CLint,max x fu = 33.0 L/h, Km 4.90, Vc 44.7 L, ka
2.40 1/h, MTT 0.502 h, NN 20.3, with VH 1 L, QH 90 L/h and fu 0.2 fixed). Its
*parameters* are on disk but its *equations* are not: the paper states only
that it fitted "a previously published one-compartment model with saturation of
elimination via a liver compartment" and cites references 41 (Kengo 2023,
Antimicrob Agents Chemother 67(11):e0043123, doi:10.1128/aac.00431-23) and 42
(Chirehwa 2016, Antimicrob Agents Chemother 60(1):487-494,
doi:10.1128/aac.01830-15, which also carries an erratum,
doi:10.1128/aac.00483-16). Neither upstream paper is open access and automated
acquisition was refused by the publisher. Rather than infer the hepatic
topology from a sibling model, the rifampicin fit has been deferred pending
acquisition of those two references. This costs nothing to the three models
shipped here: rifampicin exposure is not a covariate in any of them. The paper
screened ritonavir AUC on atazanavir clearance and rifampicin AUC on
inter-visit differences and retained neither, encoding rifampicin instead as
the categorical regimen indicator used above (the control stream's
`RTVAUC_CL` theta is `FIX`ed at 0). Both screened covariates are recorded in
`covariatesDataExcluded`.

**Allometric reference fat-free mass is 42 kg, not 41 kg.** Kengo 2025 Table 2
footnote a and the supplementary control stream (`TVFFM = 42`) both give 42 kg,
while the Table S1 and Table S3 footnotes print 41 kg -- which is the cohort
median of Table 1. The control stream governs and 42 kg is used in all three
model files. The discrepancy moves clearance by `(42/41)^0.75 = 1.8%`.

**The A5231 occasion count is inferred.** The DERIVE occasion count is pinned at
eight by the control stream's `IF (OCC==1) ... IF (OCC==8)` multiplexers over
four PK visits, i.e. two dosing occasions per visit -- the dose taken at home
before the visit and the dose given at the visit, which is also what the
`SCALE_BOV` unobserved-dose device implies. No A5231 control stream was
supplied and Kengo 2025 does not state that study's occasion count, so the
three-period A5231 design was read the same way, giving six. Because all
occasions share one variance, the count only bounds how many occasions can be
simulated; it does not change the magnitude of the between-occasion
variability.

**The unobserved-dose scaling factor is carried as a covariate.** Table 2's
"scaling factor on BOV for unobserved dose" (1.63 for DERIVE atazanavir, 1.7
for A5231) is implemented through the canonical `SELFADMIN` indicator, matching
the control stream's `IF (OBS.EQ.0) THEN BOVKA = SCALE_BOV*BOVKA ... ENDIF`.
This is a variance-model role rather than the relative-bioavailability role of
that covariate's founding example, and it is documented as such in the
covariate register. All simulations here set `SELFADMIN = 0`.

**Table 1's fat-free-mass spread is narrower than its weight spread.** Table 1
gives weight 50-75 kg but fat-free mass 37.9-41.9 kg, a range too narrow to be
the full range of a cohort with that weight spread and an 88%-female
composition; it reads more like an interquartile range despite the footnote
saying "range". Nothing in the models depends on it -- the virtual cohort above
matches the *median* of 41.0 kg, which is the quantity the allometric reference
is anchored to -- but a user reproducing Table 1 exactly should be aware of it.

**The study identifier is printed two ways.** Kengo 2025 writes the external
cohort as "ACTG A5213" in Methods 2.2 and Results 3.1 and as "A5231" in
Results 3.6, Table S2 and Table S3. The model file uses A5231, the identifier
attached to the parameter table it encodes.

**Ritonavir's ka is printed with the wrong unit.** Table 2's row header reads
"Absorption rate constant, ka (/L)". A first-order absorption rate constant has
units of reciprocal time; the model file uses 1.02 1/h. The same header appears
in Table S1 for rifampicin.

**Between-subject variability is carried on clearance only.** The control stream
fixes to zero every other between-subject eta (`$OMEGA BLOCK(1) FIX 0` for Vc,
ka, F, Vp, Q and MTT), and also the between-occasion eta on clearance and both
effect-compartment etas. The models reproduce that exactly; users should not
read the absence of an eta on volume as an omission.

**Published target attainment is asserted qualitatively, not to the percent.**
The percentages in Results 3.7 come from a 1225-individual reference cohort
drawn from previous African HIV/TB studies (Kengo 2025 reference 40) whose
typical individual weighs 61 kg, not from the DERIVE cohort. Only the DERIVE
demographics are tabulated in the article, so the virtual cohort is built on
those. The reproduction turns out to be close -- within about two percentage
points on every cell -- but the assertions are still written on the paper's
qualitative conclusions, because a 200-participant arm carries a Monte Carlo
standard error of a few percent and `rxode2::rxSetSeed()` fixes the draw within
an rxode2 version but not across versions, so a tighter bound would be a
CI-fragile assertion rather than a stronger check. The percent-level precision
in this vignette comes instead from the closed-form structural identities
above, which are tied directly to printed Table 2 and Table S3 values and are
exact at typical values.

**No between-drug correlation is modelled.** Kengo 2025 Figures S4 and S5 show
strong correlation between atazanavir and ritonavir clearance and absorption
parameters, and Results 3.4 reports that a joint atazanavir-ritonavir model and
a ritonavir-inhibition model for atazanavir were both tested and rejected in
favour of the separate models with categorical regimen covariates. The two
files reproduce the authors' final separate models; simulating them together,
as this vignette does with a shared seed, does not induce the observed
correlation between the drugs.
