---
title: "Sufentanil (Khaowroongrueng 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sufentanil (Khaowroongrueng 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Khaowroongrueng_2024_sufentanil"))$reference`
* Description: `r rxode2::rxode(readModelDb("Khaowroongrueng_2024_sufentanil"))$description`
* Article: [CPT Pharmacometrics Syst Pharmacol 2024;13(10):1682-1692](https://doi.org/10.1002/psp4.13205)

# Population

Khaowroongrueng 2024 enrolled 22 adults (>19 years) undergoing cardiac surgery
with cardiopulmonary bypass (CPB) support at a single centre in Incheon, South
Korea, between May 2021 and June 2022. Table 1 reports a median (range) age of
66 (23-77) years, weight 66.4 (46.7-99.8) kg, height 169.2 (150.1-189.2) cm and
body mass index 23.9 (16.0-34.3) kg/m^2, with 14 male and 8 female
participants. The predominant procedure was valvuloplasty or valve replacement
(13 patients), followed by coronary artery bypass grafting (6), a left
ventricular assist device (1) and combination surgery (2). Median (range) CPB
duration was 2.3 (1.4-5.6) h within a median total surgical duration of
7.7 (5.3-14.8) h. Patients on renal replacement therapy or taking drugs known to
interact with sufentanil were excluded.

Of the 110 samples collected, two participants (12 observations) were excluded
for markedly high pre-dose sufentanil concentrations and two further samples were
removed as suspected sampling errors, so the final model was fit to
**20 participants with 96 observations** (Results, "Structural model"). The
paper does not re-tabulate demographics for the 20 analysed patients, so the
ranges above are the Table 1 values for the 22 enrolled.

Age, height, weight, body mass index, sex, CYP3A4\*1G genotype (rs2242480;
13 GG / 6 GA / 3 AA), fluid replacement volumes, and the time-varying laboratory
parameters and inflammatory cytokines that changed significantly during CPB
(total protein, IL-6, IL-8, TNF-alpha) were all screened. Sex on clearance
entered the full model but was dropped at backward elimination, so **the CPB
phase effects are the only covariates retained in the final model**.

The same information is available programmatically via the model's `population`
metadata:

```{r population}
pop <- rxode2::rxode(readModelDb("Khaowroongrueng_2024_sufentanil"))$population
str(pop[c("species", "n_subjects", "age_median", "weight_median")])
```

# Source trace

Every `ini()` entry carries an in-file comment naming its source location in
`inst/modeldb/specificDrugs/Khaowroongrueng_2024_sufentanil.R`. They are
collected here for review. All final estimates come from Table 2, "Final model /
Estimate" column.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL) | 51.8 L/h (RSE 16.4%) | Table 2 |
| `lvc` (V1) | 72.5 L (RSE 12.1%) | Table 2 |
| `lq` (Q) | 55.2 L/h (RSE 17.2%) | Table 2 |
| `lvp` (V2) | 390 L (RSE 43.8%) | Table 2 |
| `e_cpb_cl` | 2.80 (RSE 14.4%) | Table 2, row "CL CPB&warming" |
| `e_cpb_vc` | 2.74 (RSE 18.5%) | Table 2, row "V1 CPB" |
| `etalcl` | 24.2 %CV -> omega^2 = log(CV^2 + 1) = 0.056921 | Table 2, row "IIV CL (CV, %)"; log-normal IIV per Methods, "Structural model" |
| (no IIV on V1 / Q / V2) | 0 FIX | Table 2, rows "IIV V1 / Q / V2 (CV, %)" |
| `propSd` | 0.253 | Table 2, row "RSV sigma_prop (%)" = 25.3% |
| Two-compartment ODEs, first-order elimination | n/a | Methods, "Structural model"; Results, "Structural model" |
| Multiplicative per-phase covariate form | n/a | Equations 1-4 |
| CL effect spans CPB + warming; V1 effect is CPB only; Q and V2 effects not retained | n/a | Results, "CPB-adjusted model" |
| Post-CPB phase reverts to the pre-CPB reference | n/a | Table 2 retains only `CL CPB&warming` and `V1 CPB`; no `COV_post-CPB` |

The paper's Equations 1-4 give, for the jth model parameter,

```
pre-CPB:   P_i,j = P_pop,j * exp(eta_i,j) * 1
CPB:       P_i,j = P_pop,j * exp(eta_i,j) * COV_CPB
warming:   P_i,j = P_pop,j * exp(eta_i,j) * COV_warming
post-CPB:  P_i,j = P_pop,j * exp(eta_i,j) * COV_post-CPB
```

that is, a plain multiplicative fold-change on the typical value within each
phase. Because the phase indicators are binary and mutually exclusive, the
model encodes this as `e_cpb_cl^(CPB_ON + CPB_REWARM)` on clearance and
`e_cpb_vc^CPB_ON` on the central volume, which reproduces the equations above
exactly. Note that the central volume returns to its baseline at the **start of
rewarming**, while clearance stays elevated until **separation from CPB**.

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the simulation scenario the authors specify in Methods, "Simulations":

* Cardiac surgery lasting 7.5 h, comprising 1.5 h of CPB and a 1 h warming phase.
* CPB commences 2.5 h after the first sufentanil dose.
* Surgery completes within 2.5 h of CPB weaning.
* Concentration-time profiles simulated for 10 h.

That fixes the phase timeline as pre-CPB [0, 2.5), CPB [2.5, 4.0), rewarming
[4.0, 5.0) and post-CPB [5.0, 10].

The recommended regimen from Results, "Simulations" is a 30 ug loading dose, a
40 ug/h maintenance infusion halted after 6 h, and an additional 60 ug IV bolus
at CPB commencement. Sufentanil was given intravenously, so every dose goes to
the `central` compartment.

A second, counterfactual arm holds both phase indicators at 0 throughout. It
receives an identical dosing history and exists only to isolate the CPB effect;
it is not a scenario the paper reports.

```{r cohort}
set.seed(20240101)

n_per_arm <- 200L

t_cpb_start <- 2.5
t_rewarm    <- 4.0
t_cpb_end   <- 5.0
t_surg_end  <- 7.5
t_inf_stop  <- 6.0

# Observation grid: a regular grid plus tightly-bracketed times at each phase
# boundary so the discontinuities in V1 and CL can be measured exactly.
eps <- 1e-5
obs_times <- sort(unique(c(
  seq(0, 10, by = 0.05),
  t_surg_end,
  c(t_cpb_start, t_rewarm, t_cpb_end, t_inf_stop) - eps,
  c(t_cpb_start, t_rewarm, t_cpb_end, t_inf_stop)
)))

# Helper: one arm as a self-contained data frame of event records.
# `on_cpb` selects whether the CPB phase indicators are exercised.
# `id_offset` keeps IDs disjoint across arms -- duplicate IDs silently
# collapse into a single subject receiving the summed dose.
make_arm <- function(n, arm, on_cpb, id_offset = 0L) {
  ids <- id_offset + seq_len(n)

  doses <- tidyr::expand_grid(
    id = ids,
    tibble::tibble(
      time = c(0, 0, t_cpb_start),
      amt  = c(30, 40 * t_inf_stop, 60),      # loading bolus, 6 h infusion, CPB bolus
      rate = c(0, 40, 0)                       # 40 ug/h infusion; 0 = bolus
    )
  ) |>
    dplyr::mutate(evid = 1L, cmt = "central")

  obs <- tidyr::expand_grid(id = ids, time = obs_times) |>
    dplyr::mutate(amt = NA_real_, rate = 0, evid = 0L, cmt = "central")

  dplyr::bind_rows(doses, obs) |>
    dplyr::mutate(
      arm        = arm,
      # Time-varying CPB phase indicators, mutually exclusive by construction.
      CPB_ON     = as.integer(on_cpb & time >= t_cpb_start & time < t_rewarm),
      CPB_REWARM = as.integer(on_cpb & time >= t_rewarm    & time < t_cpb_end)
    ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  make_arm(n_per_arm, arm = "CPB surgery",    on_cpb = TRUE,  id_offset =         0L),
  make_arm(n_per_arm, arm = "No CPB (counterfactual)", on_cpb = FALSE, id_offset = n_per_arm)
) |>
  as.data.frame()

# Disjoint-ID guard, and confirm the indicators never overlap.
# Every id must belong to exactly one arm: a shared id across arms would
# silently merge two subjects into one receiving the summed dose history.
stopifnot(
  all(tapply(events$arm, events$id, dplyr::n_distinct) == 1L),
  all(events$CPB_ON + events$CPB_REWARM <= 1L),
  dplyr::n_distinct(events$id) == 2L * n_per_arm,
  # The counterfactual arm must never switch either indicator on.
  all(events$CPB_ON[events$arm != "CPB surgery"] == 0L),
  all(events$CPB_REWARM[events$arm != "CPB surgery"] == 0L)
)
```

# Simulation

Inter-individual variability sits on clearance only (24.2 %CV); Table 2 reports
the V1, Q and V2 IIV terms as `0 FIX`, so those parameters carry no
between-subject variability.

```{r simulate}
mod <- readModelDb("Khaowroongrueng_2024_sufentanil")

sim <- rxode2::rxSolve(mod, events = events, keep = c("arm", "CPB_ON", "CPB_REWARM")) |>
  as.data.frame()

# `Cc` is the individual prediction; `sim` additionally carries the
# proportional residual error. Figure 3 of the paper is a mean simulated
# concentration-time profile, and the proportional error is mean-zero, so the
# mean of `Cc` is the right comparator.
dplyr::glimpse(sim[, c("id", "time", "arm", "cl", "vc", "central", "Cc")])
```

A deterministic typical-value solve is used for the exact structural checks
below. Rather than `zeroRe()` (which mutates shared model state), the etas are
supplied as data columns with `omega = NA`.

```{r simulate-typical}
ev_typ <- events |>
  dplyr::filter(id %in% c(1L, n_per_arm + 1L)) |>
  dplyr::mutate(etalcl = 0)

sim_typ <- rxode2::rxSolve(mod, events = ev_typ, omega = NA,
                           keep = c("arm", "CPB_ON", "CPB_REWARM")) |>
  as.data.frame()
```

# Structural checks against the published covariate model

The per-phase fold-changes are exact multiplicative statements, so they can be
gated exactly rather than eyeballed. In the CPB arm the typical-value solve must
show clearance at `51.8 * 2.80` L/h throughout CPB **and** rewarming, and the
central volume at `72.5 * 2.74` L during CPB only.

```{r structural-gates}
typ_cpb <- sim_typ |> dplyr::filter(arm == "CPB surgery", time > 0)

phase_of <- function(time) {
  dplyr::case_when(
    time <  t_cpb_start ~ "pre-CPB",
    time <  t_rewarm    ~ "CPB",
    time <  t_cpb_end   ~ "rewarming",
    TRUE                ~ "post-CPB"
  )
}

param_by_phase <- typ_cpb |>
  dplyr::mutate(phase = phase_of(time)) |>
  dplyr::group_by(phase) |>
  dplyr::summarise(cl = unique(round(cl, 6)), vc = unique(round(vc, 6)),
                   .groups = "drop") |>
  dplyr::mutate(phase = factor(phase,
    levels = c("pre-CPB", "CPB", "rewarming", "post-CPB"))) |>
  dplyr::arrange(phase)

param_by_phase |>
  dplyr::mutate(
    `CL fold-change` = round(cl / 51.8, 3),
    `V1 fold-change` = round(vc / 72.5, 3)
  ) |>
  dplyr::rename("Phase" = phase, "CL (L/h)" = cl, "V1 (L)" = vc) |>
  knitr::kable(caption = paste(
    "Typical-value CL and V1 by CPB phase. Expected fold-changes:",
    "CL 2.80 during CPB and rewarming, V1 2.74 during CPB only",
    "(Khaowroongrueng 2024 Table 2)."
  ))

expected <- tibble::tibble(
  phase = factor(c("pre-CPB", "CPB", "rewarming", "post-CPB"),
                 levels = c("pre-CPB", "CPB", "rewarming", "post-CPB")),
  cl    = 51.8 * c(1, 2.80, 2.80, 1),
  vc    = 72.5 * c(1, 2.74, 1,    1)
)

stopifnot(
  isTRUE(all.equal(param_by_phase$cl, expected$cl, tolerance = 1e-6)),
  isTRUE(all.equal(param_by_phase$vc, expected$vc, tolerance = 1e-6))
)
```

The V1 discontinuity is also visible in the concentration itself. No dose is
given at the start of rewarming and the amount in the central compartment is
continuous there, so the concentration must jump upward by exactly the 2.74-fold
V1 ratio as V1 returns to baseline.

```{r hemodilution-jump}
before <- typ_cpb |> dplyr::filter(abs(time - (t_rewarm - eps)) < 1e-9)
after  <- typ_cpb |> dplyr::filter(abs(time -  t_rewarm)        < 1e-9)

jump <- tibble::tibble(
  `Quantity`                 = c("central (ug)", "V1 (L)", "Cc (ng/mL)"),
  `Just before rewarming`    = c(before$central, before$vc, before$Cc),
  `At start of rewarming`    = c(after$central,  after$vc,  after$Cc),
  `Ratio (after / before)`   = c(after$central / before$central,
                                after$vc      / before$vc,
                                after$Cc      / before$Cc)
)

knitr::kable(jump, digits = 4, caption = paste(
  "Amount is continuous across the CPB-to-rewarming boundary while V1 reverts,",
  "so Cc rises by exactly the published 2.74-fold V1 ratio."
))

stopifnot(
  # Amount continuous across the boundary.
  isTRUE(all.equal(after$central, before$central, tolerance = 1e-4)),
  # Concentration jumps by exactly the published V1 fold-change.
  isTRUE(all.equal(after$Cc / before$Cc, 2.74, tolerance = 1e-3))
)
```

# Replicate published figures

```{r figure-3, fig.width = 7, fig.height = 4.5}
# Replicates Figure 3 of Khaowroongrueng 2024: mean simulated
# concentration-time profile for the recommended regimen (30 ug loading dose,
# 40 ug/h maintenance infusion, additional 60 ug IV bolus at CPB initiation),
# with the paper's target-concentration reference lines.
fig3 <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(
    Mean = mean(Cc),
    Q05  = quantile(Cc, 0.05),
    Q95  = quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot(fig3, aes(time, Mean, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.18, colour = NA) +
  geom_line(linewidth = 0.7) +
  annotate("rect", xmin = 0, xmax = 10, ymin = 0.3, ymax = 0.6,
           alpha = 0.08, fill = "black") +
  geom_hline(yintercept = c(0.3, 0.6), linetype = "dotted") +
  geom_hline(yintercept = 0.15, linetype = "dotted", colour = "red") +
  geom_vline(xintercept = c(t_cpb_start, t_rewarm, t_cpb_end, t_surg_end),
             linetype = "dashed", alpha = 0.35) +
  labs(
    x = "Time (h)", y = "Sufentanil concentration (ng/mL)",
    colour = NULL, fill = NULL,
    title = "Figure 3 - recommended regimen, mean simulated profile",
    caption = paste(
      "Replicates Figure 3 of Khaowroongrueng 2024. Ribbon = 5th-95th",
      "percentile. Dotted black lines: 0.3-0.6 ng/mL target for adequate",
      "anaesthesia. Dotted red line: 0.15 ng/mL end-of-surgery target.",
      "Dashed verticals: CPB start (2.5 h), rewarming (4 h), CPB end (5 h),",
      "end of surgery (7.5 h)."
    )
  ) +
  theme_bw() +
  theme(legend.position = "top", plot.caption = element_text(hjust = 0))
```

The CPB arm shows the two features the paper describes: an abrupt drop in
concentration at CPB commencement despite the 60 ug bolus (V1 increases
2.74-fold by hemodilution), and a faster decline through CPB and rewarming
(clearance increased 2.80-fold). The counterfactual no-CPB arm, which receives
the same doses, stays well above the anaesthesia band through the bypass window
-- which is the paper's argument that a regimen derived from non-CPB patients is
"not suitable" for cardiac surgery with CPB.

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

# Guarantee a time = 0 record per subject to anchor AUC from 0.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, arm, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

# Doses: the loading bolus, the 6 h infusion, and the CPB bolus. The loading
# bolus and the start of the infusion coincide at t = 0, and PKNCA requires a
# single dose record per subject and time, so the two coincident events are
# pooled into one record (270 ug delivered from t = 0, the longer of the two
# durations). None of the NCA parameters requested below is dose-normalised, so
# the pooling does not affect any reported value -- the dose object here only
# supplies the dosing times that anchor the interval.
dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::mutate(duration = ifelse(rate > 0, amt / rate, 0)) |>
  dplyr::group_by(id, arm, time) |>
  dplyr::summarise(amt = sum(amt), duration = max(duration), .groups = "drop")

stopifnot(!anyDuplicated(dose_df[, c("id", "arm", "time")]))

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id, duration = "duration")

# The surgical window [0, 7.5 h] is the interval the paper's targets refer to.
intervals <- data.frame(
  start   = 0,
  end     = t_surg_end,
  cmax    = TRUE,
  tmax    = TRUE,
  auclast = TRUE,
  cav     = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_summary <- as.data.frame(nca_res) |>
  dplyr::group_by(arm, PPTESTCD) |>
  dplyr::summarise(Median = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = Median)

nca_summary |>
  dplyr::rename(
    "Arm"                     = arm,
    "Cmax (ng/mL)"            = cmax,
    "Tmax (h)"                = tmax,
    "AUC0-7.5h (ng*h/mL)"     = auclast,
    "Cav 0-7.5h (ng/mL)"      = cav
  ) |>
  knitr::kable(digits = 3, caption = paste(
    "Median simulated NCA parameters over the 7.5 h surgical window.",
    "Khaowroongrueng 2024 reports no NCA table, so these are not a",
    "reproduction of published values."
  ))
```

## Comparison against published target concentrations

Khaowroongrueng 2024 reports no NCA table, so there are no published Cmax /
AUC / half-life values to compare against and `ncaComparisonTable()` is not
applicable. What the paper does report quantitatively about its simulation are
the target concentrations that drove the dose recommendation (Methods,
"Simulations"; Results, "Simulations"; Figure 3). Those are compared below.

```{r target-comparison}
cc_at <- function(t, arm_name = "CPB surgery") {
  sim |>
    dplyr::filter(arm == arm_name, abs(time - t) < 1e-9, !is.na(Cc)) |>
    dplyr::pull(Cc) |>
    mean()
}

# Mean concentration over the maintenance window (post-loading, pre-CPB and
# the post-CPB remainder of surgery), where the 0.3-0.6 ng/mL anaesthesia
# target applies.
maint <- sim |>
  dplyr::filter(arm == "CPB surgery", !is.na(Cc),
                time >= 0.5, time <= t_surg_end) |>
  dplyr::pull(Cc)

comparison <- tibble::tribble(
  ~Milestone,                                  ~`Paper's target`,        ~Simulated,
  "Just before CPB (2.5 h)",                   "at least 0.5-0.7 ng/mL", cc_at(t_cpb_start - eps),
  "During CPB (3.0 h)",                        "(not targeted)",         cc_at(3.0),
  "End of surgery (7.5 h)",                    "approx. 0.15 ng/mL",     cc_at(t_surg_end),
  "Mean over surgery (0.5-7.5 h)",             "0.3-0.6 ng/mL",          mean(maint)
)

comparison |>
  dplyr::mutate(Simulated = round(Simulated, 3)) |>
  knitr::kable(caption = paste(
    "Simulated mean concentrations at the milestones",
    "Khaowroongrueng 2024 used to justify the recommended regimen",
    "(Methods and Results, 'Simulations'; Figure 3)."
  ))

# The end-of-surgery target is the paper's most specific quantitative claim
# about this regimen: halting the infusion at 6 h "aligns with a target
# concentration of 0.15 ng/mL at the conclusion of the surgery".
stopifnot(
  cc_at(t_surg_end) > 0.12, cc_at(t_surg_end) < 0.19,
  # Mean exposure across the surgical window sits inside the anaesthesia band.
  mean(maint) > 0.3, mean(maint) < 0.6
)
```

The end-of-surgery concentration (0.165 ng/mL) and the mean over the surgical
window (0.373 ng/mL) both land where the paper says they should. The pre-CPB
value does not: at 0.430 ng/mL it sits below the "at least 0.5-0.7 ng/mL"
figure quoted in Methods, "Simulations".

Methods ties that figure to two doses jointly: "doses of sufentanil were varied
to propose a loading dose (25-40 ug) and an additional IV bolus dose (50-80 ug)
corresponding to a target concentration of at least 0.5-0.7 ng/mL for pre-CPB
control". The band itself is carried over from the literature (the paper's
reference 25, Hudson et al.), not estimated here. Sweeping both doses across the
ranges the paper explored locates where the band is reachable.

```{r dose-sweep}
sweep_grid <- tidyr::expand_grid(loading = c(25, 30, 40),
                                 bolus   = c(50, 60, 80)) |>
  dplyr::mutate(i = dplyr::row_number(),
                arm = paste0(loading, " ug loading + ", bolus, " ug bolus"))

ev_sweep <- lapply(seq_len(nrow(sweep_grid)), function(k) {
  g <- sweep_grid[k, ]
  make_arm(1L, arm = g$arm, on_cpb = TRUE, id_offset = 1000L * g$i) |>
    dplyr::mutate(amt = dplyr::case_when(
      # The loading bolus is the t = 0 dose record with rate == 0.
      evid == 1L & time == 0 & rate == 0 ~ g$loading,
      # The additional bolus is the dose record at CPB commencement.
      evid == 1L & time == t_cpb_start   ~ g$bolus,
      TRUE                               ~ amt
    ))
}) |>
  dplyr::bind_rows() |>
  dplyr::mutate(etalcl = 0) |>
  as.data.frame()

sweep_sim <- rxode2::rxSolve(mod, events = ev_sweep, omega = NA, keep = "arm") |>
  as.data.frame()

sweep <- sweep_sim |>
  dplyr::filter(abs(time - (t_cpb_start - eps)) < 1e-9 |
                abs(time -  t_cpb_start)        < 1e-9) |>
  dplyr::mutate(when = ifelse(time < t_cpb_start, "before", "after")) |>
  dplyr::select(arm, when, Cc) |>
  tidyr::pivot_wider(names_from = when, values_from = Cc) |>
  dplyr::transmute(
    `Loading + additional bolus`      = arm,
    `Cc just before CPB (ng/mL)`      = round(before, 3),
    `Cc just after CPB bolus (ng/mL)` = round(after, 3),
    `Either >= 0.5 ng/mL`             = ifelse(pmax(before, after) >= 0.5,
                                               "yes", "no")
  )

knitr::kable(sweep, caption = paste(
  "Typical-value concentrations at CPB commencement across the loading-dose",
  "and additional-bolus ranges Khaowroongrueng 2024 explored",
  "(Methods, 'Simulations')."
))
```

The sweep shows the pre-CPB concentration is almost insensitive to the loading
dose: by 2.5 h the loading bolus has largely distributed and cleared (V2 = 390 L,
CL = 51.8 L/h), so the concentration is set by the 40 ug/h infusion, which has
not yet reached steady state (`40 / 51.8` = 0.772 ng/mL). It is the **additional
bolus at CPB commencement** that lifts the concentration into the 0.5-0.7 band,
and only at the top of the explored 50-80 ug range -- the recommended 60 ug
bolus gives 0.456 ng/mL immediately after dosing.

So the recommended regimen does not attain the pre-CPB band under this model, and
no combination in the explored grid attains it *before* CPB. Two readings are
consistent with the paper: either the 0.5-0.7 ng/mL criterion was applied to the
post-bolus concentration at CPB initiation (where 80 ug reaches it), or the
recommendation deliberately trades it off, as Results implies by choosing 30 ug
"considering the target concentrations at various CPB phases". The paper does not
tabulate its own simulated concentrations, and Figure 3 is a raster image, so
this cannot be settled from the sources on disk. It is recorded as an open
discrepancy rather than resolved by adjusting any parameter.

# Assumptions and deviations

* **No supplementary material on disk.** Tables S1-S3 and Figures S1-S3 are not
  available for this extraction. Every final parameter value is in main-text
  Table 2, so nothing in the model file depends on the supplement. The
  supplement carries the model-development trail (Table S1), the clinical
  laboratory / cytokine data (Table S2) and a literature comparison of Vd and
  Vdss (Table S3) -- none of which enters the model.
* **Post-CPB phase collapses to the pre-CPB reference.** Equation 4 admits a
  distinct `COV_post-CPB` multiplier, but Table 2 retains only
  `CL CPB&warming` and `V1 CPB`, so no post-CPB effect was carried into the
  final model. The model therefore has no `CPB_POST` indicator.
* **CPB phase decomposed into two binary indicators.** The paper stores the
  phase as a single 0/1/2/3 column. Per the standing convention of decomposing
  categorical covariates rather than testing levels inside `model()`, this is
  encoded as the canonical `CPB_ON` (phase 1) and `CPB_REWARM` (phase 2), both
  newly registered in `inst/references/covariate-columns.md` alongside this
  extraction. The clearance effect uses the sum of the two indicators because
  the paper estimated a single typical value spanning both windows.
* **IIV variance scale.** Table 2 reports IIV as %CV and Methods states a
  log-normal IIV distribution, so `omega^2 = log(CV^2 + 1) = 0.056921` for the
  24.2 %CV clearance term. Under the alternative NONMEM convention
  (`omega^2 = CV^2`) the variance would be 0.058564; the two differ by 3% in
  variance and are indistinguishable at the reported 3 significant figures.
* **No IIV on V1, Q or V2.** Table 2 reports these as `0 FIX`, so no
  `etalvc` / `etalq` / `etalvp` terms are included. The Discussion attributes
  this to the small sample size limiting IIV estimation.
* **No IIV on the CPB effects.** The paper estimates the phase fold-changes as
  typical values only.
* **Units.** Doses are in ug and volumes in L, so `Cc = central / vc` is in
  ug/L, which is identically ng/mL -- the paper's reporting unit. No scaling
  factor is needed. `checkModelConventions()` emits an informational note that
  the dosing prefix (`ug`) and the concentration numerator (`ng`) differ in
  magnitude; that heuristic does not account for the L-versus-mL denominator,
  and the relationship is exact here.
* **Time-varying V1 produces a discontinuous concentration.** The amount in the
  central compartment is continuous across a phase boundary while V1 steps, so
  `Cc` jumps. This is the standard consequence of a time-varying volume in a
  NONMEM ADVAN3/TRANS4-style parameterisation and reproduces the paper's
  Equations 1-4 faithfully; it is not a physiological claim about instantaneous
  redistribution.
* **Simulation scenario, not observed data.** The observed concentrations are
  not public. The virtual cohort follows the phase timeline and dosing the
  authors specify for their own simulation (Methods, "Simulations"), using
  200 subjects per arm rather than the paper's 1000 replicates; the mean
  profile is insensitive to that difference at 24 %CV on clearance alone.
* **OPEN DISCREPANCY -- the pre-CPB "0.5-0.7 ng/mL" target is not attained.**
  The simulated mean concentration just before CPB under the recommended
  regimen is 0.430 ng/mL (typical value 0.422), against the "at least
  0.5-0.7 ng/mL for pre-CPB control" quoted in Methods, "Simulations". The dose
  sweep above shows this is not a loading-dose effect: across the paper's entire
  explored 25-40 ug loading range the pre-CPB concentration moves only from
  0.417 to 0.430 ng/mL, because by 2.5 h the bolus has distributed into the
  390 L peripheral compartment and been cleared at 51.8 L/h, leaving the not-yet
  steady-state 40 ug/h infusion to set the level. Only the additional bolus at
  CPB commencement reaches the band, and only at the top of its explored 50-80 ug
  range. The paper does not tabulate its simulated concentrations and Figure 3 is
  a raster image, so this cannot be settled from the sources on disk. **No
  parameter was adjusted to close the gap**; every value remains as published in
  Table 2. The two structural gates that *are* exactly checkable (the 2.80-fold
  CL and 2.74-fold V1 changes) match to 1e-6, and the paper's two specific
  claims about the recommended regimen (0.15 ng/mL at end of surgery, the
  0.3-0.6 ng/mL anaesthesia band) are both reproduced, which localises the
  discrepancy to the dose-selection narrative rather than to the PK model.
* **The counterfactual no-CPB arm is not a published scenario.** It is included
  only to isolate the CPB effect and to give the fold-change gates a paired
  reference.
* **Weight and other demographics are not in the model.** No allometric or
  demographic covariate was retained, so the virtual cohort carries no weight
  column; the loading dose is the paper's recommended flat 30 ug rather than a
  per-kg dose.
* **`half.life` is not reported.** Over the 7.5 h surgical window the profile
  never reaches a clean terminal phase (the infusion stops at 6 h), so a
  half-life estimate would describe distribution rather than terminal
  elimination and is omitted rather than reported misleadingly.
