---
title: "Sufentanil (Khaowroongrueng 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sufentanil (Khaowroongrueng 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Khaowroongrueng_2024_sufentanil"))$reference`
* Description: `r rxode2::rxode(readModelDb("Khaowroongrueng_2024_sufentanil"))$description`
* Article: [CPT Pharmacometrics Syst Pharmacol 2024;13(10):1682-1692](https://doi.org/10.1002/psp4.13205)

# Population

Khaowroongrueng 2024 enrolled 22 adults (>19 years) undergoing cardiac surgery
with cardiopulmonary bypass (CPB) support at a single centre in Incheon, South
Korea, between May 2021 and June 2022. Table 1 reports a median (range) age of
66 (23-77) years, weight 66.4 (46.7-99.8) kg, height 169.2 (150.1-189.2) cm and
body mass index 23.9 (16.0-34.3) kg/m^2, with 14 male and 8 female
participants. The predominant procedure was valvuloplasty or valve replacement
(13 patients), followed by coronary artery bypass grafting (6), a left
ventricular assist device (1) and combination surgery (2). Median (range) CPB
duration was 2.3 (1.4-5.6) h within a median total surgical duration of
7.7 (5.3-14.8) h. Patients on renal replacement therapy or taking drugs known to
interact with sufentanil were excluded.

Of the 110 samples collected, two participants (12 observations) were excluded
for markedly high pre-dose sufentanil concentrations and two further samples were
removed as suspected sampling errors, so the final model was fit to
**20 participants with 96 observations** (Results, "Structural model"). The
paper does not re-tabulate demographics for the 20 analysed patients, so the
ranges above are the Table 1 values for the 22 enrolled.

Age, height, weight, body mass index, sex, CYP3A4\*1G genotype (rs2242480;
13 GG / 6 GA / 3 AA), fluid replacement volumes, and the time-varying laboratory
parameters and inflammatory cytokines that changed significantly during CPB
(total protein, IL-6, IL-8, TNF-alpha) were all screened. Sex on clearance
entered the full model but was dropped at backward elimination, so **the CPB
phase effects are the only covariates retained in the final model**.

The same information is available programmatically via the model's `population`
metadata:

```{r population}
pop <- rxode2::rxode(readModelDb("Khaowroongrueng_2024_sufentanil"))$population
str(pop[c("species", "n_subjects", "age_median", "weight_median")])
```

# Source trace

Every `ini()` entry carries an in-file comment naming its source location in
`inst/modeldb/specificDrugs/Khaowroongrueng_2024_sufentanil.R`. They are
collected here for review. All final estimates come from Table 2, "Final model /
Estimate" column.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL) | 51.8 L/h (RSE 16.4%) | Table 2 |
| `lvc` (V1) | 72.5 L (RSE 12.1%) | Table 2 |
| `lq` (Q) | 55.2 L/h (RSE 17.2%) | Table 2 |
| `lvp` (V2) | 390 L (RSE 43.8%) | Table 2 |
| `e_cpb_cl` | 2.80 (RSE 14.4%) | Table 2, row "CL CPB&warming" |
| `e_cpb_vc` | 2.74 (RSE 18.5%) | Table 2, row "V1 CPB" |
| `etalcl` | 24.2 %CV -> omega^2 = log(CV^2 + 1) = 0.056921 | Table 2, row "IIV CL (CV, %)"; log-normal IIV per Methods, "Structural model" |
| (no IIV on V1 / Q / V2) | 0 FIX | Table 2, rows "IIV V1 / Q / V2 (CV, %)" |
| `propSd` | 0.253 | Table 2, row "RSV sigma_prop (%)" = 25.3% |
| Two-compartment ODEs, first-order elimination | n/a | Methods, "Structural model"; Results, "Structural model" |
| Multiplicative per-phase covariate form | n/a | Equations 1-4 |
| CL effect spans CPB + warming; V1 effect is CPB only; Q and V2 effects not retained | n/a | Results, "CPB-adjusted model" |
| Post-CPB phase reverts to the pre-CPB reference | n/a | Table 2 retains only `CL CPB&warming` and `V1 CPB`; no `COV_post-CPB` |

The paper's Equations 1-4 give, for the jth model parameter,

```
pre-CPB:   P_i,j = P_pop,j * exp(eta_i,j) * 1
CPB:       P_i,j = P_pop,j * exp(eta_i,j) * COV_CPB
warming:   P_i,j = P_pop,j * exp(eta_i,j) * COV_warming
post-CPB:  P_i,j = P_pop,j * exp(eta_i,j) * COV_post-CPB
```

that is, a plain multiplicative fold-change on the typical value within each
phase. Because the phase indicators are binary and mutually exclusive, the
model encodes this as `e_cpb_cl^(CPB_ON + CPB_REWARM)` on clearance and
`e_cpb_vc^CPB_ON` on the central volume, which reproduces the equations above
exactly. Note that the central volume returns to its baseline at the **start of
rewarming**, while clearance stays elevated until **separation from CPB**.

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the simulation scenario the authors specify in Methods, "Simulations":

* Cardiac surgery lasting 7.5 h, comprising 1.5 h of CPB and a 1 h warming phase.
* CPB commences 2.5 h after the first sufentanil dose.
* Surgery completes within 2.5 h of CPB weaning.
* Concentration-time profiles simulated for 10 h.

That fixes the phase timeline as pre-CPB [0, 2.5), CPB [2.5, 4.0), rewarming
[4.0, 5.0) and post-CPB [5.0, 10].

The recommended regimen from Results, "Simulations" is a 30 ug loading dose, a
40 ug/h maintenance infusion halted after 6 h, and an additional 60 ug IV bolus
at CPB commencement. Sufentanil was given intravenously, so every dose goes to
the `central` compartment.

A second, counterfactual arm holds both phase indicators at 0 throughout. It
receives an identical dosing history and exists only to isolate the CPB effect;
it is not a scenario the paper reports.

```{r cohort}
set.seed(20240101)

n_per_arm <- 200L

t_cpb_start <- 2.5
t_rewarm    <- 4.0
t_cpb_end   <- 5.0
t_surg_end  <- 7.5
t_inf_stop  <- 6.0

# Observation grid: a regular grid plus tightly-bracketed times at each phase
# boundary so the discontinuities in V1 and CL can be measured exactly.
eps <- 1e-5
obs_times <- sort(unique(c(
  seq(0, 10, by = 0.05),
  t_surg_end,
  c(t_cpb_start, t_rewarm, t_cpb_end, t_inf_stop) - eps,
  c(t_cpb_start, t_rewarm, t_cpb_end, t_inf_stop)
)))

# Helper: one arm as a self-contained data frame of event records.
# `on_cpb` selects whether the CPB phase indicators are exercised.
# `id_offset` keeps IDs disjoint across arms -- duplicate IDs silently
# collapse into a single subject receiving the summed dose.
make_arm <- function(n, arm, on_cpb, id_offset = 0L) {
  ids <- id_offset + seq_len(n)

  doses <- tidyr::expand_grid(
    id = ids,
    tibble::tibble(
      time = c(0, 0, t_cpb_start),
      amt  = c(30, 40 * t_inf_stop, 60),      # loading bolus, 6 h infusion, CPB bolus
      rate = c(0, 40, 0)                       # 40 ug/h infusion; 0 = bolus
    )
  ) |>
    dplyr::mutate(evid = 1L, cmt = "central")

  obs <- tidyr::expand_grid(id = ids, time = obs_times) |>
    dplyr::mutate(amt = NA_real_, rate = 0, evid = 0L, cmt = "central")

  dplyr::bind_rows(doses, obs) |>
    dplyr::mutate(
      arm        = arm,
      # Time-varying CPB phase indicators, mutually exclusive by construction.
      CPB_ON     = as.integer(on_cpb & time >= t_cpb_start & time < t_rewarm),
      CPB_REWARM = as.integer(on_cpb & time >= t_rewarm    & time < t_cpb_end)
    ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  make_arm(n_per_arm, arm = "CPB surgery",    on_cpb = TRUE,  id_offset =         0L),
  make_arm(n_per_arm, arm = "No CPB (counterfactual)", on_cpb = FALSE, id_offset = n_per_arm)
) |>
  as.data.frame()

# Disjoint-ID guard, and confirm the indicators never overlap.
# Every id must belong to exactly one arm: a shared id across arms would
# silently merge two subjects into one receiving the summed dose history.
stopifnot(
  all(tapply(events$arm, events$id, dplyr::n_distinct) == 1L),
  all(events$CPB_ON + events$CPB_REWARM <= 1L),
  dplyr::n_distinct(events$id) == 2L * n_per_arm,
  # The counterfactual arm must never switch either indicator on.
  all(events$CPB_ON[events$arm != "CPB surgery"] == 0L),
  all(events$CPB_REWARM[events$arm != "CPB surgery"] == 0L)
)
```

# Simulation

Inter-individual variability sits on clearance only (24.2 %CV); Table 2 reports
the V1, Q and V2 IIV terms as `0 FIX`, so those parameters carry no
between-subject variability.

```{r simulate}
mod <- readModelDb("Khaowroongrueng_2024_sufentanil")

sim <- rxode2::rxSolve(mod, events = events, keep = c("arm", "CPB_ON", "CPB_REWARM")) |>
  as.data.frame()

# `Cc` is the individual prediction; `sim` additionally carries the
# proportional residual error. Figure 3 of the paper is a mean simulated
# concentration-time profile, and the proportional error is mean-zero, so the
# mean of `Cc` is the right comparator.
dplyr::glimpse(sim[, c("id", "time", "arm", "cl", "vc", "central", "Cc")])
```

A deterministic typical-value solve is used for the exact structural checks
below. Rather than `zeroRe()` (which mutates shared model state), the etas are
supplied as data columns with `omega = NA`.

```{r simulate-typical}
ev_typ <- events |>
  dplyr::filter(id %in% c(1L, n_per_arm + 1L)) |>
  dplyr::mutate(etalcl = 0)

sim_typ <- rxode2::rxSolve(mod, events = ev_typ, omega = NA,
                           keep = c("arm", "CPB_ON", "CPB_REWARM")) |>
  as.data.frame()
```

# Structural checks against the published covariate model

The per-phase fold-changes are exact multiplicative statements, so they can be
gated exactly rather than eyeballed. In the CPB arm the typical-value solve must
show clearance at `51.8 * 2.80` L/h throughout CPB **and** rewarming, and the
central volume at `72.5 * 2.74` L during CPB only.

```{r structural-gates}
typ_cpb <- sim_typ |> dplyr::filter(arm == "CPB surgery", time > 0)

phase_of <- function(time) {
  dplyr::case_when(
    time <  t_cpb_start ~ "pre-CPB",
    time <  t_rewarm    ~ "CPB",
    time <  t_cpb_end   ~ "rewarming",
    TRUE                ~ "post-CPB"
  )
}

param_by_phase <- typ_cpb |>
  dplyr::mutate(phase = phase_of(time)) |>
  dplyr::group_by(phase) |>
  dplyr::summarise(cl = unique(round(cl, 6)), vc = unique(round(vc, 6)),
                   .groups = "drop") |>
  dplyr::mutate(phase = factor(phase,
    levels = c("pre-CPB", "CPB", "rewarming", "post-CPB"))) |>
  dplyr::arrange(phase)

param_by_phase |>
  dplyr::mutate(
    `CL fold-change` = round(cl / 51.8, 3),
    `V1 fold-change` = round(vc / 72.5, 3)
  ) |>
  dplyr::rename("Phase" = phase, "CL (L/h)" = cl, "V1 (L)" = vc) |>
  knitr::kable(caption = paste(
    "Typical-value CL and V1 by CPB phase. Expected fold-changes:",
    "CL 2.80 during CPB and rewarming, V1 2.74 during CPB only",
    "(Khaowroongrueng 2024 Table 2)."
  ))

expected <- tibble::tibble(
  phase = factor(c("pre-CPB", "CPB", "rewarming", "post-CPB"),
                 levels = c("pre-CPB", "CPB", "rewarming", "post-CPB")),
  cl    = 51.8 * c(1, 2.80, 2.80, 1),
  vc    = 72.5 * c(1, 2.74, 1,    1)
)

stopifnot(
  isTRUE(all.equal(param_by_phase$cl, expected$cl, tolerance = 1e-6)),
  isTRUE(all.equal(param_by_phase$vc, expected$vc, tolerance = 1e-6))
)
```

The V1 discontinuity is also visible in the concentration itself. No dose is
given at the start of rewarming and the amount in the central compartment is
continuous there, so the concentration must jump upward by exactly the 2.74-fold
V1 ratio as V1 returns to baseline.

```{r hemodilution-jump}
before <- typ_cpb |> dplyr::filter(abs(time - (t_rewarm - eps)) < 1e-9)
after  <- typ_cpb |> dplyr::filter(abs(time -  t_rewarm)        < 1e-9)

jump <- tibble::tibble(
  `Quantity`                 = c("central (ug)", "V1 (L)", "Cc (ng/mL)"),
  `Just before rewarming`    = c(before$central, before$vc, before$Cc),
  `At start of rewarming`    = c(after$central,  after$vc,  after$Cc),
  `Ratio (after / before)`   = c(after$central / before$central,
                                after$vc      / before$vc,
                                after$Cc      / before$Cc)
)

knitr::kable(jump, digits = 4, caption = paste(
  "Amount is continuous across the CPB-to-rewarming boundary while V1 reverts,",
  "so Cc rises by exactly the published 2.74-fold V1 ratio."
))

stopifnot(
  # Amount continuous across the boundary.
  isTRUE(all.equal(after$central, before$central, tolerance = 1e-4)),
  # Concentration jumps by exactly the published V1 fold-change.
  isTRUE(all.equal(after$Cc / before$Cc, 2.74, tolerance = 1e-3))
)
```

# Replicate published figures

```{r figure-3, fig.width = 7, fig.height = 4.5}
# Replicates Figure 3 of Khaowroongrueng 2024: mean simulated
# concentration-time profile for the recommended regimen (30 ug loading dose,
# 40 ug/h maintenance infusion, additional 60 ug IV bolus at CPB initiation),
# with the paper's target-concentration reference lines.
fig3 <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(
    Mean = mean(Cc),
    Q05  = quantile(Cc, 0.05),
    Q95  = quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot(fig3, aes(time, Mean, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.18, colour = NA) +
  geom_line(linewidth = 0.7) +
  annotate("rect", xmin = 0, xmax = 10, ymin = 0.3, ymax = 0.6,
           alpha = 0.08, fill = "black") +
  geom_hline(yintercept = c(0.3, 0.6), linetype = "dotted") +
  geom_hline(yintercept = 0.15, linetype = "dotted", colour = "red") +
  geom_vline(xintercept = c(t_cpb_start, t_rewarm, t_cpb_end, t_surg_end),
             linetype = "dashed", alpha = 0.35) +
  labs(
    x = "Time (h)", y = "Sufentanil concentration (ng/mL)",
    colour = NULL, fill = NULL,
    title = "Figure 3 - recommended regimen, mean simulated profile",
    caption = paste(
      "Replicates Figure 3 of Khaowroongrueng 2024. Ribbon = 5th-95th",
      "percentile. Dotted black lines: 0.3-0.6 ng/mL target for adequate",
      "anaesthesia. Dotted red line: 0.15 ng/mL end-of-surgery target.",
      "Dashed verticals: CPB start (2.5 h), rewarming (4 h), CPB end (5 h),",
      "end of surgery (7.5 h)."
    )
  ) +
  theme_bw() +
  theme(legend.position = "top", plot.caption = element_text(hjust = 0))
```

The two CPB effects act in opposite directions at commencement, so the net step
is computed rather than asserted. Results states the extra 60 ug bolus was given
"to account for the hemodilution effect", i.e. it is meant to offset the
2.74-fold rise in V1; whether it under- or over-compensates is a quantitative
question about this model.

```{r cpb-onset}
mean_cc <- function(t, arm_name) {
  mean(sim$Cc[sim$arm == arm_name & abs(sim$time - t) < 1e-9], na.rm = TRUE)
}

# Mean central amount just before CPB, used to build the "dilution alone"
# counterfactual: what the concentration would have been at commencement had
# the 60 ug bolus not been given and only the 2.74-fold V1 step applied.
amt_pre  <- mean(sim$central[sim$arm == "CPB surgery" &
                             abs(sim$time - (t_cpb_start - eps)) < 1e-9])
dilution_only <- amt_pre / (72.5 * 2.74)

cc_pre    <- mean_cc(t_cpb_start - eps, "CPB surgery")
cc_onset  <- mean_cc(t_cpb_start,       "CPB surgery")
onset_pct <- 100 * (cc_onset / cc_pre - 1)

onset <- tibble::tibble(
  Quantity = c(
    "CPB arm, just before CPB (2.5 h)",
    "CPB arm, at commencement (2.74-fold V1 step + 60 ug bolus)",
    "Counterfactual: 2.74-fold dilution alone, no bolus",
    "CPB arm, end of CPB (4.0 h, before V1 reverts)",
    "CPB arm, start of rewarming (V1 reverts to baseline)",
    "No-CPB arm, at CPB start (2.5 h)",
    "No-CPB arm, at CPB end (5.0 h)"
  ),
  `Mean Cc (ng/mL)` = c(
    cc_pre, cc_onset, dilution_only,
    mean_cc(t_rewarm - eps, "CPB surgery"),
    mean_cc(t_rewarm,       "CPB surgery"),
    mean_cc(t_cpb_start,    "No CPB (counterfactual)"),
    mean_cc(t_cpb_end,      "No CPB (counterfactual)")
  )
)

knitr::kable(onset, digits = 3, caption = paste(
  "Mean simulated concentrations across the bypass window. The",
  "dilution-alone row is a counterfactual, not a simulated arm: it applies",
  "the 2.74-fold V1 step to the observed pre-CPB amount without the 60 ug",
  "bolus."
))

# When the counterfactual arm's mean profile drops back below the 0.6 ng/mL
# anaesthesia ceiling, expressed as a fraction of the way through the bypass
# window. Quoted in the narrative below.
noc <- sim |>
  dplyr::filter(arm == "No CPB (counterfactual)", !is.na(Cc),
                time >= t_cpb_start, time <= t_cpb_end) |>
  dplyr::group_by(time) |>
  dplyr::summarise(mean_cc = mean(Cc), .groups = "drop")
t_cross  <- min(noc$time[noc$mean_cc < 0.6])
frac_win <- (t_cross - t_cpb_start) / (t_cpb_end - t_cpb_start)

# Monte-Carlo precision of the mean profile at this cohort size: the largest
# standard error of the mean across the observation grid, as a percentage of the
# mean at that time. Quoted in "Assumptions and deviations" below.
sem_max <- sim |>
  dplyr::filter(arm == "CPB surgery", !is.na(Cc), time > 0) |>
  dplyr::group_by(time) |>
  dplyr::summarise(sem_pct = 100 * sd(Cc) / sqrt(dplyr::n()) / mean(Cc),
                   .groups = "drop") |>
  dplyr::pull(sem_pct) |>
  max()

stopifnot(
  # The bolus more than offsets hemodilution: the net step at commencement is UP.
  onset_pct > 0,
  # Dilution alone would have cut the concentration by more than half.
  dilution_only < 0.5 * cc_pre,
  # Clearance is 2.80-fold higher through CPB, so the decline is faster: the
  # concentration must fall below its pre-CPB level by the end of the phase.
  mean_cc(t_rewarm - eps, "CPB surgery") < cc_pre,
  # The counterfactual arm never drops below the CPB arm during the window.
  mean_cc(t_cpb_start, "No CPB (counterfactual)") > cc_onset
)
```

The 60 ug bolus therefore **over**-compensates: the mean concentration steps
*up* by `r sprintf("%.1f%%", onset_pct)` at commencement
(`r sprintf("%.3f", cc_pre)` to `r sprintf("%.3f", cc_onset)` ng/mL), whereas the
2.74-fold dilution acting alone would have taken it to
`r sprintf("%.3f", dilution_only)` ng/mL. What the CPB arm does show is the
second effect, the 2.80-fold clearance increase: the concentration falls from
`r sprintf("%.3f", cc_onset)` to
`r sprintf("%.3f", mean_cc(t_rewarm - eps, "CPB surgery"))` ng/mL over the 1.5 h
bypass phase, well below its pre-CPB level. V1 then reverts at the start of
rewarming and the concentration jumps to
`r sprintf("%.3f", mean_cc(t_rewarm, "CPB surgery"))` ng/mL, the highest mean in
this arm -- which is why the NCA table above reports Tmax at 4.0 h rather than at
a dosing time.

The counterfactual no-CPB arm, receiving an identical dosing history, runs above
the CPB arm throughout the bypass window
(`r sprintf("%.3f", mean_cc(t_cpb_start, "No CPB (counterfactual)"))` ng/mL at
CPB start, declining to
`r sprintf("%.3f", mean_cc(t_cpb_end, "No CPB (counterfactual)"))` ng/mL by CPB
end) -- it starts well above the 0.6 ng/mL anaesthesia ceiling and does not drop
back into the 0.3-0.6 band until `r sprintf("%.2f", t_cross)` h, about
`r sprintf("%.0f%%", 100 * frac_win)` of the way through the window. This is the
model-side
counterpart of the paper's Discussion claim that "using pharmacokinetic
parameters obtained from patients undergoing general anesthesia for sufentanil
TCI during cardiac surgery with CPB is not suitable": ignoring the phase effects
overstates exposure during bypass.

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

# Guarantee a time = 0 record per subject to anchor AUC from 0.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, arm, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

# Doses: the loading bolus, the 6 h infusion, and the CPB bolus. The loading
# bolus and the start of the infusion coincide at t = 0, and PKNCA requires a
# single dose record per subject and time, so the two coincident events are
# pooled into one record (270 ug delivered from t = 0, the longer of the two
# durations). None of the NCA parameters requested below is dose-normalised, so
# the pooling does not affect any reported value -- the dose object here only
# supplies the dosing times that anchor the interval.
dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::mutate(duration = ifelse(rate > 0, amt / rate, 0)) |>
  dplyr::group_by(id, arm, time) |>
  dplyr::summarise(amt = sum(amt), duration = max(duration), .groups = "drop")

stopifnot(!anyDuplicated(dose_df[, c("id", "arm", "time")]))

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id, duration = "duration")

# The surgical window [0, 7.5 h] is the interval the paper's targets refer to.
intervals <- data.frame(
  start   = 0,
  end     = t_surg_end,
  cmax    = TRUE,
  tmax    = TRUE,
  auclast = TRUE,
  cav     = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_summary <- as.data.frame(nca_res) |>
  dplyr::group_by(arm, PPTESTCD) |>
  dplyr::summarise(Median = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = Median)

nca_summary |>
  dplyr::rename(
    "Arm"                     = arm,
    "Cmax (ng/mL)"            = cmax,
    "Tmax (h)"                = tmax,
    "AUC0-7.5h (ng*h/mL)"     = auclast,
    "Cav 0-7.5h (ng/mL)"      = cav
  ) |>
  knitr::kable(digits = 3, caption = paste(
    "Median simulated NCA parameters over the 7.5 h surgical window.",
    "Khaowroongrueng 2024 reports no NCA table, so these are not a",
    "reproduction of published values."
  ))
```

## Comparison against published target concentrations

Khaowroongrueng 2024 reports no NCA table, so there are no published Cmax /
AUC / half-life values to compare against and `ncaComparisonTable()` is not
applicable. What the paper does report quantitatively about its simulation are
the target concentrations that drove the dose recommendation (Methods,
"Simulations"; Results, "Simulations"; Figure 3). Those are compared below.

```{r target-comparison}
cc_at <- function(t, arm_name = "CPB surgery") {
  sim |>
    dplyr::filter(arm == arm_name, abs(time - t) < 1e-9, !is.na(Cc)) |>
    dplyr::pull(Cc) |>
    mean()
}

# Mean concentration over the maintenance window (post-loading, pre-CPB and
# the post-CPB remainder of surgery), where the 0.3-0.6 ng/mL anaesthesia
# target applies.
maint <- sim |>
  dplyr::filter(arm == "CPB surgery", !is.na(Cc),
                time >= 0.5, time <= t_surg_end) |>
  dplyr::pull(Cc)

comparison <- tibble::tribble(
  ~Milestone,                                  ~`Paper's target`,        ~Simulated,
  "Just before CPB (2.5 h)",                   "at least 0.5-0.7 ng/mL", cc_at(t_cpb_start - eps),
  "During CPB (3.0 h)",                        "(not targeted)",         cc_at(3.0),
  "End of surgery (7.5 h)",                    "approx. 0.15 ng/mL",     cc_at(t_surg_end),
  "Mean over surgery (0.5-7.5 h)",             "0.3-0.6 ng/mL",          mean(maint)
)

comparison |>
  dplyr::mutate(Simulated = round(Simulated, 3)) |>
  knitr::kable(caption = paste(
    "Simulated mean concentrations at the milestones",
    "Khaowroongrueng 2024 used to justify the recommended regimen",
    "(Methods and Results, 'Simulations'; Figure 3)."
  ))

# The end-of-surgery target is the paper's most specific quantitative claim
# about this regimen: Results recommends the infusion "be halted after 6 h,
# aligning with a target concentration of 0.15 ng/mL at the conclusion of the
# surgery", on the assumption that surgery concludes within 1.5 h of cessation.
stopifnot(
  cc_at(t_surg_end) > 0.12, cc_at(t_surg_end) < 0.19,
  # Mean exposure across the surgical window sits inside the anaesthesia band.
  mean(maint) > 0.3, mean(maint) < 0.6
)
```

The end-of-surgery concentration (`r sprintf("%.3f", cc_at(t_surg_end))` ng/mL)
and the mean over the surgical window (`r sprintf("%.3f", mean(maint))` ng/mL)
both land where the paper says they should. The pre-CPB value does not: at
`r sprintf("%.3f", cc_at(t_cpb_start - eps))` ng/mL it sits below the "at least
0.5-0.7 ng/mL" figure quoted in Methods, "Simulations".

Methods ties that figure to three quantities jointly: "doses of sufentanil were
varied to propose a loading dose (25-40 ug) and an additional IV bolus dose
(50-80 ug) corresponding to a target concentration of at least 0.5-0.7 ng/mL for
pre-CPB control, and an infusion rate (30-45 ug/h) corresponding to a target
concentration of 0.3-0.6 ng/mL for adequate anesthesia". The band itself is
carried over from the literature (the paper's reference 25, Hudson et al.), not
estimated here. The sweep below therefore spans the full grid the paper explored
-- all three ranges, not just the two bolus doses -- so that the question "is the
band reachable anywhere in the explored space?" is actually tested rather than
assumed.

```{r dose-sweep}
sweep_grid <- tidyr::expand_grid(loading  = c(25, 30, 40),
                                 bolus    = c(50, 60, 80),
                                 infusion = c(30, 40, 45)) |>
  dplyr::mutate(i = dplyr::row_number(),
                arm = paste0(loading, " / ", bolus, " / ", infusion))

ev_sweep <- lapply(seq_len(nrow(sweep_grid)), function(k) {
  g <- sweep_grid[k, ]
  make_arm(1L, arm = g$arm, on_cpb = TRUE, id_offset = 1000L * g$i) |>
    dplyr::mutate(
      # `amt` is computed before `rate` is overwritten, so the rate == 0 /
      # rate > 0 tests below still see the original infusion flag.
      amt = dplyr::case_when(
        # The loading bolus is the t = 0 dose record with rate == 0.
        evid == 1L & time == 0 & rate == 0 ~ g$loading,
        # The infusion is the t = 0 dose record with rate > 0; its total amount
        # must track the rate so the duration stays at t_inf_stop.
        evid == 1L & time == 0 & rate >  0 ~ g$infusion * t_inf_stop,
        # The additional bolus is the dose record at CPB commencement.
        evid == 1L & time == t_cpb_start   ~ g$bolus,
        TRUE                               ~ amt
      ),
      rate = ifelse(evid == 1L & time == 0 & rate > 0, g$infusion, rate)
    )
}) |>
  dplyr::bind_rows() |>
  dplyr::mutate(etalcl = 0) |>
  as.data.frame()

sweep_sim <- rxode2::rxSolve(mod, events = ev_sweep, omega = NA, keep = "arm") |>
  as.data.frame()

sweep <- sweep_sim |>
  dplyr::filter(abs(time - (t_cpb_start - eps)) < 1e-9 |
                abs(time -  t_cpb_start)        < 1e-9) |>
  dplyr::mutate(when = ifelse(time < t_cpb_start, "before", "after")) |>
  dplyr::select(arm, when, Cc) |>
  tidyr::pivot_wider(names_from = when, values_from = Cc) |>
  dplyr::left_join(sweep_grid, by = "arm") |>
  dplyr::arrange(infusion, bolus, loading)

# Summary quantities the narrative below quotes, so the prose cannot drift from
# the table.
sweep_max_before <- max(sweep$before)
sweep_n_before   <- sum(sweep$before >= 0.5)
sweep_n_after    <- sum(sweep$after  >= 0.5)
# Spread attributable to the loading dose alone, holding bolus and infusion at
# the recommended 60 ug / 40 ug/h.
sweep_rec <- sweep |> dplyr::filter(bolus == 60, infusion == 40)

sweep |>
  dplyr::transmute(
    `Loading (ug)`                    = loading,
    `CPB bolus (ug)`                  = bolus,
    `Infusion (ug/h)`                 = infusion,
    `Cc just before CPB (ng/mL)`      = round(before, 3),
    `Cc just after CPB bolus (ng/mL)` = round(after, 3),
    `Either >= 0.5 ng/mL`             = ifelse(pmax(before, after) >= 0.5,
                                               "yes", "no")
  ) |>
  knitr::kable(caption = paste(
    "Typical-value concentrations at CPB commencement across the full",
    "loading-dose, additional-bolus and infusion-rate grid Khaowroongrueng",
    "2024 explored (Methods, 'Simulations')."
  ))

stopifnot(
  # The claim in the narrative: the pre-CPB band is not reached anywhere in the
  # explored grid before the additional bolus is given.
  sweep_n_before == 0L,
  # It IS reached post-bolus somewhere in the grid.
  sweep_n_after > 0L
)
```

The sweep shows the pre-CPB concentration is almost insensitive to the loading
dose: by 2.5 h the loading bolus has largely distributed and cleared (V2 = 390 L,
CL = 51.8 L/h), so the level is set mostly by the infusion, which has not yet
reached steady state (at the recommended 40 ug/h the asymptote is
`40 / 51.8` = 0.772 ng/mL). Holding the bolus and infusion at the recommended
60 ug and 40 ug/h, the entire 25-40 ug loading range moves the pre-CPB
concentration only from `r sprintf("%.3f", min(sweep_rec$before))` to
`r sprintf("%.3f", max(sweep_rec$before))` ng/mL. It is the **additional bolus at
CPB commencement** that lifts the concentration into the 0.5-0.7 band, and only
at the top of the explored 50-80 ug range -- the recommended 60 ug bolus gives
`r sprintf("%.3f", sweep_rec$after[sweep_rec$loading == 30])` ng/mL immediately
after dosing.

So the recommended regimen does not attain the pre-CPB band under this model, and
**no combination anywhere in the paper's explored grid attains it before CPB**:
across all `r nrow(sweep)` loading / bolus / infusion combinations the highest
pre-CPB concentration is `r sprintf("%.3f", sweep_max_before)` ng/mL, reached at
the top of the loading and infusion ranges (40 ug loading, 45 ug/h infusion --
the additional bolus is given at CPB commencement and so cannot affect the
pre-CPB value), still short of 0.5. Post-bolus the band is attainable, in
`r sweep_n_after` of `r nrow(sweep)` combinations. Two readings are consistent
with the paper: either the 0.5-0.7 ng/mL criterion was applied to the post-bolus
concentration at CPB initiation, or the recommendation deliberately trades it off,
as Results implies by choosing 30 ug "considering the target concentrations at
various CPB phases". The paper does not tabulate its own simulated
concentrations, and Figure 3 is a raster image, so this cannot be settled from
the sources on disk. It is recorded as an open discrepancy rather than resolved
by adjusting any parameter.

# Assumptions and deviations

* **No supplementary material on disk.** Tables S1-S3 and Figures S1-S3 are not
  available for this extraction. Every final parameter value is in main-text
  Table 2, so nothing in the model file depends on the supplement. The
  supplement carries the model-development trail (Table S1), the clinical
  laboratory / cytokine data (Table S2) and a literature comparison of Vd and
  Vdss (Table S3) -- none of which enters the model.
* **Post-CPB phase collapses to the pre-CPB reference.** Equation 4 admits a
  distinct `COV_post-CPB` multiplier, but Table 2 retains only
  `CL CPB&warming` and `V1 CPB`, so no post-CPB effect was carried into the
  final model. The model therefore has no `CPB_POST` indicator.
* **CPB phase decomposed into two binary indicators.** The paper stores the
  phase as a single 0/1/2/3 column. Per the standing convention of decomposing
  categorical covariates rather than testing levels inside `model()`, this is
  encoded as the canonical `CPB_ON` (phase 1) and `CPB_REWARM` (phase 2), both
  newly registered in `inst/references/covariate-columns.md` alongside this
  extraction. The clearance effect uses the sum of the two indicators because
  the paper estimated a single typical value spanning both windows.
* **IIV variance scale.** Table 2 reports IIV as %CV and Methods states a
  log-normal IIV distribution, so `omega^2 = log(CV^2 + 1) = 0.056921` for the
  24.2 %CV clearance term. Under the alternative NONMEM convention
  (`omega^2 = CV^2`) the variance would be 0.058564; the two differ by 3% in
  variance and are indistinguishable at the reported 3 significant figures.
* **No IIV on V1, Q or V2.** Table 2 reports these as `0 FIX`, so no
  `etalvc` / `etalq` / `etalvp` terms are included. The Discussion attributes
  this to the small sample size limiting IIV estimation.
* **No IIV on the CPB effects.** The paper estimates the phase fold-changes as
  typical values only.
* **Units.** Doses are in ug and volumes in L, so `Cc = central / vc` is in
  ug/L, which is identically ng/mL -- the paper's reporting unit. No scaling
  factor is needed. `checkModelConventions()` emits an informational note that
  the dosing prefix (`ug`) and the concentration numerator (`ng`) differ in
  magnitude; that heuristic does not account for the L-versus-mL denominator,
  and the relationship is exact here.
* **Time-varying V1 produces a discontinuous concentration.** The amount in the
  central compartment is continuous across a phase boundary while V1 steps, so
  `Cc` jumps. This is the standard consequence of a time-varying volume in a
  NONMEM ADVAN3/TRANS4-style parameterisation and reproduces the paper's
  Equations 1-4 faithfully; it is not a physiological claim about instantaneous
  redistribution.
* **Simulation scenario, not observed data.** The observed concentrations are
  not public. The virtual cohort follows the phase timeline and dosing the
  authors specify for their own simulation (Methods, "Simulations"), using
  200 subjects per arm rather than the paper's 1000 replicates. With IIV on
  clearance only (24.2 %CV), the Monte-Carlo standard error of the mean profile
  at 200 per arm peaks at `r sprintf("%.2f%%", sem_max)` of the mean across the
  observation grid, so the mean profile the figure and tables report is well
  determined at this cohort size.
* **OPEN DISCREPANCY -- the pre-CPB "0.5-0.7 ng/mL" target is not attained.**
  The simulated mean concentration just before CPB under the recommended
  regimen is `r sprintf("%.3f", cc_at(t_cpb_start - eps))` ng/mL (typical value
  `r sprintf("%.3f", sweep_rec$before[sweep_rec$loading == 30])`), against the
  "at least 0.5-0.7 ng/mL for pre-CPB control" quoted in Methods,
  "Simulations". The dose sweep above shows this is not a loading-dose effect:
  holding the bolus and infusion at their recommended values, the paper's entire
  25-40 ug loading range moves the pre-CPB concentration only from
  `r sprintf("%.3f", min(sweep_rec$before))` to
  `r sprintf("%.3f", max(sweep_rec$before))` ng/mL, because by 2.5 h the bolus
  has distributed into the 390 L peripheral compartment and been cleared at
  51.8 L/h, leaving the not-yet steady-state infusion to set the level. Sweeping
  the paper's full three-way grid (loading, additional bolus and infusion rate)
  does not close the gap either: the highest pre-CPB concentration anywhere in
  the `r nrow(sweep)` combinations is
  `r sprintf("%.3f", sweep_max_before)` ng/mL. Only the additional bolus at
  CPB commencement reaches the band, and only at the top of its explored 50-80 ug
  range. The paper does not tabulate its simulated concentrations and Figure 3 is
  a raster image, so this cannot be settled from the sources on disk. **No
  parameter was adjusted to close the gap**; every value remains as published in
  Table 2. The two structural gates that *are* exactly checkable (the 2.80-fold
  CL and 2.74-fold V1 changes) match to 1e-6, and the paper's two specific
  claims about the recommended regimen (0.15 ng/mL at end of surgery, the
  0.3-0.6 ng/mL anaesthesia band) are both reproduced, which localises the
  discrepancy to the dose-selection narrative rather than to the PK model.
* **The counterfactual no-CPB arm is not a published scenario.** It is included
  only to isolate the CPB effect and to give the fold-change gates a paired
  reference.
* **Weight and other demographics are not in the model.** No allometric or
  demographic covariate was retained, so the virtual cohort carries no weight
  column; the loading dose is the paper's recommended flat 30 ug rather than a
  per-kg dose.
* **`half.life` is not reported.** Over the 7.5 h surgical window the profile
  never reaches a clean terminal phase (the infusion stops at 6 h), so a
  half-life estimate would describe distribution rather than terminal
  elimination and is omitted rather than reported misleadingly.
