---
title: "Proguanil (Khwarg 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Proguanil (Khwarg 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Khwarg_2024_proguanil"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Article: <https://doi.org/10.1111/cts.70103>\n")
```

`r ui$description`

This is a **semi-physiologic joint parent-metabolite** model: rather than
attaching the transporter genotype to a systemic clearance term, Khwarg 2024
places a well-stirred liver compartment between the absorption depot and the
central compartment and lets the *SLC22A1* (OCT1) genotype act on the
OCT1-mediated hepatocyte uptake inside the hepatic extraction ratio. That is
the point of the paper -- as its Discussion puts it, "directly applying a
transporter genotype as a covariate to systemic clearance is not
straightforward, considering that hepatic uptake and clearance do not have a
proportional relationship."

# Population

The model was fit to the PK data of the CYP2C19-mediated tegoprazan/proguanil
drug-drug-interaction study NCT04568772 (Khwarg 2024 Methods "Data"), using
160 proguanil and 159 cycloguanil plasma concentrations from 16 healthy Korean
subjects sampled over 48 h after a single oral atovaquone/proguanil 250/100 mg
dose (i.e. 100 mg proguanil hydrochloride), given alone or with tegoprazan
50 mg. The esomeprazole and vonoprazan arms of the parent study were excluded
because of known interactions with proguanil.

All subjects were male, aged 19-50 y at eligibility (cohort mean 34.31 y,
SD 7.38), with mean height 172.85 cm (SD 7.37), mean body weight 76.12 kg
(SD 9.82) and mean BMI 25.52 kg/m^2 (SD 3.14) (Khwarg 2024 Results
"Demographics"; per-genotype demographics in Table S2). Every subject was a
CYP2C19 normal metabolizer (*1/*1), which is what isolates the *SLC22A1*
effect from CYP2C19 phenotype as a confounder.

Thirteen of the 16 subjects consented to genotyping. Of the six assayed
decreased-function *SLC22A1* SNPs, only 1022C>T (rs2282143) was polymorphic:
9 subjects were CC homozygous wild type and 4 were CT heterozygous; no TT
homozygotes were observed (Khwarg 2024 Table 1). The three ungenotyped
subjects were assigned by a mixture model with Hardy-Weinberg probabilities
fixed at 0.7 (CC) / 0.3 (CT), and all three were classified CC.

The same information is available programmatically via the model's
`population` metadata (`readModelDb("Khwarg_2024_proguanil")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Khwarg_2024_proguanil.R`. The
table below collects them in one place for review.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lka` (KA) | 0.101 1/h | Table 3, Proguanil (RSE 6.0%) |
| `lcl` (CL/F) | 33.5 L/h | Table 3, Proguanil (RSE 16.4%) |
| `lvc` (Vc/F) | 58.4 L | Table 3, Proguanil (RSE 24.1%) |
| `lq` (Q/F) | 41.4 L/h | Table 3, Proguanil (RSE 7.2%) |
| `lvp` (Vp/F) | 1400 L | Table 3, Proguanil (RSE 26.0%) |
| `lclint` (CLint) | 78.3 L/h | Table 3, Proguanil (RSE 18.6%) |
| `e_snp_oct1_clint` (FUP) | 0.416 | Table 3, Proguanil (RSE 54.1%) |
| `lktr_cycloguanil` (mKT) | 1.03 1/h | Table 3, Cycloguanil (RSE 5.8%) |
| `lcl_cycloguanil` (CLM/F) | 97.0 L/h | Table 3, Cycloguanil (RSE 13.8%) |
| `etalcl` | 31.7 %CV -> 0.095755 | Table 3 IIV CL/F; footnote CV = sqrt(exp(omega^2)-1)*100 |
| `etalvc` | 87.2 %CV -> 0.565532 | Table 3 IIV Vc/F |
| `etalclint` | 53.7 %CV -> 0.253377 | Table 3 IIV CLint |
| `etalktr_cycloguanil` | 15.5 %CV -> 0.023741 | Table 3 IIV mKT |
| `propSd` | 0.16 | Table 3, Proguanil proportional error (RSE 9.1%) |
| `propSd_cycloguanil` | 0.124 | Table 3, Cycloguanil proportional error (RSE 12.7%) |
| `addSd_cycloguanil` | 0.372 ug/L | Table 3, Cycloguanil additive error (RSE 18.5%) |
| `fu_fix` (fu) | 0.25 | Methods "Structural model" (refs 7, 24) |
| `bpr_fix` (CB/CP) | 2.7 | Methods "Structural model" (refs 7, 24) |
| `hct_fix` (hematocrit) | 0.45 | Methods "Structural model" / Equation 7 |
| `lvh_70` (VH at 70 kg) | 1 L | Equation 5 |
| `lqbh_70` (QBH at 70 kg) | 90 L/h | Equation 6 |
| `e_wt_vh` | 1 | Equation 5, `(body weight / 70 kg)` with no exponent |
| `e_wt_qbh` | 0.75 | Equation 6, `(body weight / 70 kg)^0.75` |
| `fub = fu / (CB/CP)` | n/a | Equation 3 |
| `eh` well-stirred with `FUP^GENO` | n/a | Equation 8 (Equation 1 is its GENO = 0 case) |
| `CLBH = eh * QBH` | n/a | Equation 2 |
| `CLH = CLBH * (CB/CP)` | n/a | Equation 4 |
| `QH = QBH * (1 - hematocrit)` | n/a | Equation 7 |
| ODE topology (all 8 states, all flows) | n/a | Figure 2 (model schematic) |
| Molar-mass ratio 288.17 / 290.19 | 0.993004 | Results "Structural model" |

# The fixed physiologic sub-model

Equations 3, 5, 6 and 7 are pure physiology and were not estimated. This
chunk reproduces them independently of the model file, so the numbers the
ODEs consume are visible and checkable.

```{r physiology}
wt_mean <- 76.12 # kg, Khwarg 2024 Results "Demographics"
fu <- 0.25
bpr <- 2.7
hct <- 0.45

fub <- fu / bpr # Eq 3
vh <- 1 * (wt_mean / 70)^1 # Eq 5, L
qbh <- 90 * (wt_mean / 70)^0.75 # Eq 6, L/h
qh <- qbh * (1 - hct) # Eq 7, L/h

tibble::tibble(
  Quantity = c(
    "fub = fu / (CB/CP)", "VH (L)", "QBH (L/h)", "QH (L/h)"
  ),
  Equation = c("3", "5", "6", "7"),
  Value = round(c(fub, vh, qbh, qh), 4)
) |>
  knitr::kable(caption = "Fixed physiologic constants at the cohort mean body weight.")
```

The genotype effect enters Equation 8 as an exponent on FUP, then propagates
to the hepatic blood clearance (Equation 2) and the hepatic plasma clearance
(Equation 4). Because the dose is absorbed into the liver, the fraction of an
absorbed dose that survives first pass is
`QH * (1 - EH) / (QH * (1 - EH) + CLH)`.

```{r extraction}
clint <- 78.3
fup <- 0.416

extraction <- lapply(0:1, function(geno) {
  uptake <- fub * clint * fup^geno # Eq 8 numerator term
  eh <- uptake / (qbh + uptake) # Eq 8
  cl_bh <- eh * qbh # Eq 2
  cl_h <- cl_bh * bpr # Eq 4
  back <- qh * (1 - eh) # Figure 2, liver -> central
  tibble::tibble(
    Genotype = c("CC (GENO = 0)", "CT (GENO = 1)")[geno + 1],
    EH = eh, `CLBH (L/h)` = cl_bh, `CLH (L/h)` = cl_h,
    `First-pass surviving fraction` = back / (back + cl_h)
  )
}) |>
  dplyr::bind_rows()

extraction |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 4))) |>
  knitr::kable(caption = "Hepatic extraction by SLC22A1 1022C>T genotype (Equations 2, 4, 8).")
```

The CT heterozygote's extraction ratio is
`r round(extraction$EH[2] / extraction$EH[1], 3)`-fold that of the CC wild
type -- less drug is taken up into the hepatocyte, so more parent survives to
systemic plasma and less cycloguanil is formed. That is exactly the mechanism
the paper reports.

# Virtual cohort

Original observed data are not publicly available. Two virtual arms of 200
subjects each are built, one per *SLC22A1* 1022C>T genotype, with body weights
drawn to match the published mean and SD and truncated to the study's BMI
eligibility range at the mean height.

```{r cohort}
set.seed(20241212)

n_per_arm <- 200L
obs_times <- seq(0, 48, by = 0.5)

make_arm <- function(n, geno, label, id_offset = 0L) {
  # Body weight: mean 76.12 kg, SD 9.82 kg (Khwarg 2024 Results
  # "Demographics"), truncated to the BMI 19-30 kg/m^2 eligibility band at
  # the cohort mean height of 172.85 cm.
  ht_m <- 172.85 / 100
  wt <- pmin(pmax(rnorm(n, 76.12, 9.82), 19 * ht_m^2), 30 * ht_m^2)
  subj <- tibble::tibble(
    id = id_offset + seq_len(n),
    WT = wt,
    SNP_SLC22A1_RS2282143 = geno,
    genotype = label
  )
  dplyr::bind_rows(
    # Single oral 100 mg proguanil hydrochloride into the depot.
    subj |> dplyr::mutate(
      time = 0, amt = 100, evid = 1L,
      cmt = "depot", dvid = NA_integer_
    ),
    # Observation rows. Both model outputs are ALGEBRAIC endpoints, so the
    # observation record is addressed by `dvid` with a missing compartment;
    # rxode2 then returns every observable as a column on every output row.
    # Naming an observable in the compartment column instead would inject a
    # compartment slot for it and renumber the ODE states.
    subj |> tidyr::crossing(time = obs_times) |> dplyr::mutate(
      amt = NA_real_, evid = 0L,
      cmt = NA_character_, dvid = 1L
    )
  ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  make_arm(n_per_arm, 0, "CC", id_offset = 0L),
  make_arm(n_per_arm, 1, "CT", id_offset = n_per_arm)
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
stopifnot(nrow(dplyr::distinct(events, id)) == 2L * n_per_arm)
```

# Simulation

```{r simulate}
mod <- readModelDb("Khwarg_2024_proguanil")

# `useLinCmt = FALSE` -- rxode2's automatic ODE -> linCmt conversion corrupts
# the dvid -> cmt mapping for multi-output models of this shape.
sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep = c("WT", "SNP_SLC22A1_RS2282143", "genotype"),
  useLinCmt = FALSE,
  returnType = "data.frame"
) |>
  dplyr::filter(!is.na(time))

# Typical-value (no between-subject variability) profiles at the cohort mean
# weight, used for the Figure 1 overlay.
typical_events <- dplyr::bind_rows(
  make_arm(1L, 0, "CC", id_offset = 0L),
  make_arm(1L, 1, "CT", id_offset = 1L)
) |>
  dplyr::mutate(WT = wt_mean)

sim_typical <- rxode2::rxSolve(
  mod,
  events = typical_events,
  keep = c("genotype"),
  omega = NA,
  useLinCmt = FALSE,
  returnType = "data.frame"
) |>
  dplyr::filter(!is.na(time))
```

# Replicate published figures

## Figure 1 -- mean concentration-time profiles and metabolic ratio by genotype

Khwarg 2024 Figure 1 plots the mean proguanil (panel a) and cycloguanil
(panel b) plasma profiles and their ratio (panel c) by genotype. The cohort
means from the virtual population are overlaid with the typical-value
prediction.

```{r figure-1}
# Replicates Figure 1 of Khwarg 2024.
long_mean <- sim |>
  dplyr::group_by(genotype, time) |>
  dplyr::summarise(
    Proguanil = mean(Cc),
    Cycloguanil = mean(Cc_cycloguanil),
    `Cycloguanil / proguanil ratio` = mean(Cc_cycloguanil / Cc, na.rm = TRUE),
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(-c(genotype, time), names_to = "panel", values_to = "value") |>
  dplyr::mutate(panel = factor(panel, levels = c(
    "Proguanil", "Cycloguanil", "Cycloguanil / proguanil ratio"
  )))

long_typ <- sim_typical |>
  dplyr::group_by(genotype, time) |>
  dplyr::summarise(
    Proguanil = mean(Cc),
    Cycloguanil = mean(Cc_cycloguanil),
    `Cycloguanil / proguanil ratio` = mean(Cc_cycloguanil / Cc, na.rm = TRUE),
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(-c(genotype, time), names_to = "panel", values_to = "value") |>
  dplyr::mutate(panel = factor(panel, levels = levels(long_mean$panel)))

ggplot(long_mean, aes(time, value, colour = genotype)) +
  geom_line(linewidth = 0.9) +
  geom_line(
    data = long_typ, aes(time, value, colour = genotype),
    linetype = "dashed", linewidth = 0.5
  ) +
  facet_wrap(~panel, scales = "free_y") +
  labs(
    x = "Time (h)", y = "Plasma concentration (ug/L) or ratio",
    colour = "SLC22A1 1022C>T",
    title = "Figure 1 - mean profiles by SLC22A1 genotype",
    caption = paste(
      "Replicates Figure 1 of Khwarg 2024. Solid = virtual-cohort mean,",
      "dashed = typical value."
    )
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

The CT arm sits above the CC arm for proguanil and below it for cycloguanil,
and the ratio panel separates the two genotypes by roughly twofold -- the
qualitative result the paper reports in its Abstract and Figure 1.

## Figure 3 -- visual predictive check by genotype and analyte

Khwarg 2024 Figure 3 is a four-panel VPC (proguanil and cycloguanil, CC and
CT). The prediction intervals below are the 5th, 50th and 95th percentiles of
the simulated cohort, matching the percentiles the paper's VPC displays.

```{r figure-3}
# Replicates Figure 3 of Khwarg 2024.
sim |>
  dplyr::select(id, time, genotype, Proguanil = Cc, Cycloguanil = Cc_cycloguanil) |>
  tidyr::pivot_longer(c(Proguanil, Cycloguanil), names_to = "analyte", values_to = "conc") |>
  dplyr::group_by(analyte, genotype, time) |>
  dplyr::summarise(
    Q05 = quantile(conc, 0.05),
    Q50 = quantile(conc, 0.50),
    Q95 = quantile(conc, 0.95),
    .groups = "drop"
  ) |>
  dplyr::filter(time > 0) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(linewidth = 0.8) +
  facet_grid(analyte ~ genotype, scales = "free_y") +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Plasma concentration (ug/L)",
    title = "Figure 3 - VPC by analyte and SLC22A1 genotype",
    caption = "Replicates Figure 3 of Khwarg 2024. Band = 5th-95th percentile, line = median."
  ) +
  theme_bw()
```

# PKNCA validation

Khwarg 2024 Table 2 reports AUC~last~ over the 0-48 h sampling window,
C~max~, T~max~ and terminal half-life for each analyte by genotype. One PKNCA
block is run per output, over the matching 0-48 h interval.

```{r pknca-setup}
dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, genotype)

intervals <- data.frame(
  start = 0, end = 48,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, half.life = TRUE
)

run_nca <- function(conc_col) {
  # Only `!is.na(Cc)` -- adding `time > 0` or `Cc > 0` would drop the
  # time-zero row that PKNCA needs to anchor AUC0-*.
  d <- sim |>
    dplyr::mutate(Cc = .data[[conc_col]]) |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::select(id, time, Cc, genotype)
  # Defensive time-zero row (pre-dose concentration is 0 for an oral dose).
  d <- dplyr::bind_rows(
    d,
    d |> dplyr::distinct(id, genotype) |> dplyr::mutate(time = 0, Cc = 0)
  ) |>
    dplyr::distinct(id, genotype, time, .keep_all = TRUE) |>
    dplyr::arrange(id, genotype, time)
  stopifnot(nrow(d) > 0)
  PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(d, Cc ~ time | genotype + id),
    PKNCA::PKNCAdose(dose_df, amt ~ time | genotype + id),
    intervals = intervals
  ))
}

nca_proguanil <- run_nca("Cc")
nca_cycloguanil <- run_nca("Cc_cycloguanil")
```

## Comparison against published NCA -- proguanil

Reference values are the "Total" columns of Khwarg 2024 Table 2 (CC N = 9,
CT N = 4), which pool the proguanil-alone and proguanil + tegoprazan
occasions; tegoprazan is not a covariate in the final model, so the pooled
column is the correct comparator.

```{r nca-compare-proguanil}
published_proguanil <- tibble::tribble(
  ~genotype, ~cmax,  ~tmax, ~auclast, ~half.life,
  "CC",      78.97,  3.00,  1205.29,  15.90,
  "CT",      93.34,  2.50,  1502.00,  18.07
)

nlmixr2lib::ncaComparisonTable(
  simulated = nca_proguanil,
  reference = published_proguanil,
  by = "genotype",
  units = c(cmax = "ug/L", auclast = "h*ug/L", tmax = "h", half.life = "h"),
  tolerance_pct = 20
) |>
  knitr::kable(
    caption = paste(
      "Proguanil: simulated vs Khwarg 2024 Table 2 (Total columns).",
      "* differs from reference by >20%."
    )
  )
```

## Comparison against published NCA -- cycloguanil

```{r nca-compare-cycloguanil}
published_cycloguanil <- tibble::tribble(
  ~genotype, ~cmax,  ~tmax, ~auclast, ~half.life,
  "CC",      29.17,  6.00,  445.08,   11.24,
  "CT",      17.48,  6.00,  279.48,   13.02
)

nlmixr2lib::ncaComparisonTable(
  simulated = nca_cycloguanil,
  reference = published_cycloguanil,
  by = "genotype",
  units = c(cmax = "ug/L", auclast = "h*ug/L", tmax = "h", half.life = "h"),
  tolerance_pct = 20
) |>
  knitr::kable(
    caption = paste(
      "Cycloguanil: simulated vs Khwarg 2024 Table 2 (Total columns).",
      "* differs from reference by >20%."
    )
  )
```

## Why the half-life rows are starred

Both comparison tables star `t1/2`: PKNCA returns roughly 29 h for proguanil
against the published 15.9 h. That gap is a property of the *measurement*, not
a disagreement with the model, and this section shows why.

The published parameter set carries a large, slowly-equilibrating peripheral
compartment (Vp/F = 1400 L, RSE 26%, against Vc/F = 58.4 L), so the system's
true terminal slope is far slower than anything a 48 h sampling window can
resolve. The analytic terminal slope of the two-compartment system,
`beta = 0.5 * [(k10 + k12 + k21) - sqrt((k10 + k12 + k21)^2 - 4 * k10 * k21)]`,
gives a terminal half-life of about 44 h. A concentration-time curve truncated
at 48 h is therefore still in the transition phase throughout, and the apparent
half-life depends entirely on which points the regression uses.

```{r halflife-window}
apparent_thalf <- function(dat, col, lo, hi) {
  d <- dat[dat$time >= lo & dat$time <= hi, ]
  log(2) / -stats::coef(stats::lm(log(d[[col]]) ~ d$time))[[2]]
}

windows <- list(c(12, 48), c(24, 48), c(36, 48))

thalf_tab <- lapply(windows, function(w) {
  tibble::tibble(
    Window = sprintf("%d-%d h", w[1], w[2]),
    Proguanil = apparent_thalf(sim_typical[sim_typical$genotype == "CC", ], "Cc", w[1], w[2]),
    Cycloguanil = apparent_thalf(
      sim_typical[sim_typical$genotype == "CC", ], "Cc_cycloguanil", w[1], w[2]
    )
  )
}) |>
  dplyr::bind_rows() |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 1)))

thalf_tab |>
  dplyr::bind_rows(tibble::tibble(
    Window = "Published (Table 2, CC)", Proguanil = 15.90, Cycloguanil = 11.24
  )) |>
  knitr::kable(caption = paste(
    "Apparent terminal half-life (h) of the typical-value CC profile by",
    "regression window, against the published NCA values."
  ))
```

Khwarg 2024 sampled to 48 h with terminal observations at roughly 8, 12, 24,
36 and 48 h (Figure 1), so a conventional lambda-z fit uses a window starting
around 8-12 h. On exactly that window the model reproduces the published
half-lives closely -- proguanil
`r thalf_tab$Proguanil[thalf_tab$Window == "12-48 h"]` h against 15.90 h, and
cycloguanil `r thalf_tab$Cycloguanil[thalf_tab$Window == "24-48 h"]` h on the
24-48 h window against 11.24 h. PKNCA's automatic best-fit lambda-z, run here
on a noise-free 0.5 h grid, instead selects only the last few points, where the
curve has bent further toward its true 44 h terminal slope. The starred rows
are that instrument difference.

This is also the reason AUC~last~ agrees so much better than `t1/2`: AUC over
0-48 h is dominated by the well-resolved absorption and distribution phases,
which the model matches to within 10% for proguanil and 5% for cycloguanil in
the CC arm. A transcription error in Vp/F or Q/F would have degraded AUC too.

## Metabolic ratio and genotype contrasts

The paper's headline results are stated as genotype ratios rather than
absolute values: proguanil exposure "higher around 1.2-fold", cycloguanil
exposure "lower around 0.6-fold", and the metabolic ratio "lower around 0.5-
to 0.6-fold" in CT versus CC (Khwarg 2024 Results "Pharmacokinetics"). This
table checks each contrast, plus the absolute metabolic ratio from Table 2.

```{r genotype-contrasts}
auc_by_subject <- function(nca_res, name) {
  out <- as.data.frame(nca_res) |>
    dplyr::filter(PPTESTCD == "auclast") |>
    dplyr::select(genotype, id, PPORRES)
  stats::setNames(out, c("genotype", "id", name))
}

mr <- auc_by_subject(nca_proguanil, "auc_p") |>
  dplyr::inner_join(auc_by_subject(nca_cycloguanil, "auc_m"), by = c("genotype", "id")) |>
  dplyr::group_by(genotype) |>
  dplyr::summarise(
    auc_p = mean(auc_p), auc_m = mean(auc_m),
    mr = mean(auc_m / auc_p), .groups = "drop"
  )

get1 <- function(g, col) {
  v <- mr[[col]][mr$genotype == g]
  if (length(v) != 1L) stop("no unique row for genotype ", g)
  v
}

tibble::tibble(
  Contrast = c(
    "Proguanil AUClast, CT / CC",
    "Cycloguanil AUClast, CT / CC",
    "Metabolic ratio, CC",
    "Metabolic ratio, CT",
    "Metabolic ratio, CT / CC"
  ),
  Simulated = c(
    get1("CT", "auc_p") / get1("CC", "auc_p"),
    get1("CT", "auc_m") / get1("CC", "auc_m"),
    get1("CC", "mr"),
    get1("CT", "mr"),
    get1("CT", "mr") / get1("CC", "mr")
  ),
  Published = c(1.25, 0.63, 0.38, 0.21, 0.55),
  `Published source` = c(
    "Table 2: 1502.00 / 1205.29",
    "Table 2: 279.48 / 445.08",
    "Table 2, metabolic ratio CC",
    "Table 2, metabolic ratio CT",
    "Table 2: 0.21 / 0.38"
  )
) |>
  dplyr::mutate(
    `Ratio to published` = round(Simulated / Published, 3),
    Simulated = round(Simulated, 3)
  ) |>
  knitr::kable(caption = "Genotype contrasts: simulated vs Khwarg 2024 Table 2 and Results text.")
```

## Mass-balance gate on the metabolite

An independent closed-form check that does not use the simulation: at steady
state the cycloguanil-forming clearance seen from the central compartment is
`CLH * CH / Cc`, so the fraction of the dose converted to cycloguanil and the
resulting AUC follow from the parameters alone. Agreement with the simulated
AUC confirms the ODE topology transcribed from Figure 2.

```{r mass-balance}
cl <- 33.5
clm <- 97.0
mwr <- 288.17 / 290.19
dose <- 100

closed_form <- lapply(0:1, function(geno) {
  uptake <- fub * clint * fup^geno
  eh <- uptake / (qbh + uptake)
  cl_h <- eh * qbh * bpr
  back <- qh * (1 - eh)
  ch_cc <- qh / (back + cl_h) # liver/central concentration ratio at ss
  cl_cyc <- cl_h * ch_cc # cyclo-pathway CL from central
  cl_tot <- cl + cl_cyc
  fh <- back / (back + cl_h) # first-pass survival
  amt_m <- dose * cl_h / (back + cl_h) + dose * fh * cl_cyc / cl_tot
  tibble::tibble(
    genotype = c("CC", "CT")[geno + 1],
    `AUCinf proguanil (h*ug/L)` = fh * dose / cl_tot * 1000,
    `AUCinf cycloguanil (h*ug/L)` = amt_m * mwr / clm * 1000
  )
}) |>
  dplyr::bind_rows()

closed_form |>
  dplyr::left_join(
    mr |> dplyr::transmute(
      genotype,
      `Simulated AUClast proguanil` = round(auc_p),
      `Simulated AUClast cycloguanil` = round(auc_m)
    ),
    by = "genotype"
  ) |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 0))) |>
  knitr::kable(caption = paste(
    "Closed-form AUCinf from the parameters alone vs the simulated 0-48 h",
    "AUClast. AUCinf exceeds AUClast by the 48 h tail, as expected."
  ))
```

# Assumptions and deviations

- **Residual-error scale read as standard deviations, not variances.**
  Khwarg 2024 Table 3 reports "Proportional error" 0.16 (proguanil), and
  "Proportional error" 0.124 plus "Additive error (ug/L)" 0.372
  (cycloguanil), without stating the scale. They are encoded as SDs. The
  discriminator is the unit label on the additive row: a variance would carry
  `(ug/L)^2`, not `ug/L`. Read as SDs the proportional terms are 16% and
  12.4%, the ordinary magnitudes for a Phase-1 LC-MS/MS assay; read as
  variances they would be 40% and 35%, which is implausible for the
  well-behaved profiles in Figure 1.
- **Equations 1-8 were recovered from the PDF text layer.** The
  preprocessor's markdown conversion emitted `formula-not-decoded` for all
  eight displayed equations, so they were read from the PDF's embedded text
  (`pdftotext -layout`). The `0.75` exponent of Equation 6 in particular is
  present only in the superscript run of that text layer.
- **The mixture model is not part of the packaged model.** Khwarg 2024 used a
  two-component mixture with fixed Hardy-Weinberg probabilities (0.7 CC /
  0.3 CT) to impute the genotype of the three ungenotyped subjects during
  estimation; all three were classified CC. That is a data-imputation device
  for the original fit, not a structural feature to redistribute, so the
  packaged model takes the genotype as a supplied covariate
  (`SNP_SLC22A1_RS2282143`). Simulating a mixed population is a matter of
  sampling the covariate at the published 0.7 / 0.3 proportions.
- **Tegoprazan co-administration is not represented.** The estimation dataset
  pooled the proguanil-alone and proguanil + tegoprazan 50 mg occasions and
  the final model carries no tegoprazan term, so the comparison above uses
  Table 2's pooled "Total" columns. Table 2's per-occasion columns differ by
  well under the genotype effect.
- **No covariates are retained on any PK parameter.** Body weight, eGFR, AST
  and ALT were screened by stepwise covariate modeling and none was retained
  (Results "Covariate analysis"). Body weight nevertheless appears in the
  model because Equations 5 and 6 scale the *fixed physiologic* liver volume
  and hepatic blood flow on it; eGFR, AST and ALT are recorded in
  `covariatesDataExcluded` to preserve the screen, with no point estimates to
  encode.
- **The liver return flow carries the `(1 - EH)` factor exactly as drawn.**
  Figure 2 labels the liver-to-central arrow `QH * (1 - EH)` while separately
  drawing the metabolic arrow `CLH` out of the same compartment. Both terms
  are transcribed literally. Coding the return flow as plain `QH` instead
  changes the predicted AUCs by under 4% at these parameter values, so the
  choice is not resolvable from the reported NCA; the printed figure decides
  it.
- **Cycloguanil shares the individual central volume of proguanil**, including
  its eta, following Results "Structural model" ("assumed to be the same as
  the central volume of distribution of proguanil, due to identifiability
  issues").
- **Dosing units.** Doses are given in mg of proguanil hydrochloride and
  volumes in L, so amount/volume is mg/L; both observations are multiplied by
  1000 to reach the ug/L scale of the published concentrations and of
  `addSd_cycloguanil`.
- **Virtual-cohort covariates.** Body weight is drawn as
  `N(76.12, 9.82)` kg truncated to the BMI 19-30 kg/m^2 eligibility band at
  the cohort mean height; the paper reports the mean and SD but not the
  distributional form. All subjects are male CYP2C19 normal metabolizers, as
  in the study. Arms of 200 subjects per genotype are simulated rather than
  the study's 9 and 4, so the cohort means are stable; Table 2's SDs are
  therefore not comparable to the simulated spread.
- **T~max~ comparison is grid-limited in the source, not here.** Table 2's
  T~max~ is a median over the study's sparse sampling times, whereas the
  simulation is evaluated on a 0.5 h grid, so small T~max~ differences reflect
  sampling resolution rather than model disagreement. Proguanil T~max~ is
  starred (2.0 h simulated against a published median of 3.00 h with observed
  range 1.00-4.00 h), which is one sampling interval of the source schedule.
- **Terminal half-life is starred in both comparison tables and was
  investigated, not tuned.** The published parameter set implies an analytic
  terminal half-life near 44 h, which a 48 h sampling window cannot resolve, so
  the apparent value depends on the regression window. On the window a
  conventional lambda-z fit would use for this study's sampling schedule the
  model reproduces the published half-lives (see "Why the half-life rows are
  starred"). No parameter was adjusted; AUC~last~ agreement confirms the
  distributional parameters were transcribed correctly.
- **The metabolic-ratio contrast in the CT arm runs low.** The simulated
  CT/CC metabolic-ratio contrast is about 0.42 against the 0.5-0.6 range the
  Results text quotes, and the absolute CT metabolic ratio is 0.17 against
  Table 2's 0.21. Two features of the source explain the direction. Table 2's
  NCA covers only the 9 CC and 4 CT subjects who were actually genotyped,
  whereas the fit that produced FUP = 0.416 additionally assigned three
  ungenotyped subjects to CC via the mixture model; and FUP itself carries the
  loosest precision of any parameter in Table 3 (RSE 54.1%), so the genotype
  contrast is the least well-determined quantity in the model. The 1-sigma
  interval on FUP spans roughly 0.19-0.64, which brackets the published
  contrast comfortably.
