---
title: "Midazolam and 1-OH-midazolam during and after VA-ECMO (Kim 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Midazolam and 1-OH-midazolam during and after VA-ECMO (Kim 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Kim 2026 is a *Research Letter* that nonetheless carries an original joint
parent/metabolite population PK model with a complete parameter table, so it is
extractable. The authors fit **two independent models** to the same prospective
cohort -- one during VA-ECMO support and one after decannulation
(Supplementary Methods: "Two independent population pharmacokinetic models were
developed") -- so the paper contributes two model files and this single
vignette.

```{r load-models}
mod_ecmo <- readModelDb("Kim_2026_midazolam_ecmo")
mod_post <- readModelDb("Kim_2026_midazolam_postecmo")

ui_ecmo <- rxode2::rxode(mod_ecmo)
ui_post <- rxode2::rxode(mod_post)
```

* Citation: `r ui_ecmo$reference`
* Article: <https://doi.org/10.1097/ALN.0000000000005811>
* Supplemental methods: <https://links.lww.com/ALN/E277>
* Supplemental tables: <https://links.lww.com/ALN/E278>
* Supplemental figures: <https://links.lww.com/ALN/E279>

On-ECMO model description:

> `r ui_ecmo$description`

Post-ECMO model description:

> `r ui_post$description`

# Population

Nineteen adults receiving midazolam-based sedation during venoarterial ECMO in a
cardiovascular intensive care unit contributed to the on-ECMO analysis; eleven
of them had post-decannulation sampling and contributed to the post-ECMO
analysis. Baseline characteristics are Supplementary Table S1: median age 53.5 y
(22-87), median weight 70 kg (52.9-92), 5 of 19 female, median VA-ECMO duration
133.7 h (10.6-269). The cohort is markedly hepatically stressed (median AST
234 U/L, range 3.8-2883), which matters for a drug cleared by CYP3A4.
Indications for VA-ECMO were ST-elevation myocardial infarction (10), atrial
fibrillation (3), myocarditis (2), cardiac arrest (2), non-ST-elevation
myocardial infarction (1), and pulmonary embolism (1). Patients with severe
hepatic or renal dysfunction, or on strong CYP3A4 inhibitors, were excluded.
The study is ClinicalTrials.gov NCT02581280 (Yonsei University IRB 4-2014-0919).

Midazolam was given as a continuous infusion at a rate chosen by the treating
physician, with supplementary boluses as needed; the study did not fix a dose.

The same information is available programmatically:

```{r population}
str(ui_ecmo$population, max.level = 1)
```

# Model structure

Both models share the structure of Supplementary Fig. S1: a two-compartment
midazolam disposition whose **only** elimination pathway is metabolic clearance
to 1-OH-midazolam (`CL_MF`), feeding a two-compartment 1-OH-midazolam
disposition that is cleared by `CL_1-OH`.

```
IV infusion --> [central (MDZ)] <--Q_MDZ--> [peripheral1 (MDZ)]
                      |
                      | CL_MF   (the only midazolam elimination)
                      v
                [central_1ohm] <--Q_1-OH--> [peripheral1_1ohm]
                      |
                      | CL_1-OH
                      v
```

Because `CL_MF` is the sole parent elimination, every milligram of midazolam
cleared appears as one milligram-equivalent of 1-OH-midazolam. The paper reports
no molar conversion factor, so the transfer is encoded 1:1 in
midazolam-equivalent mass -- the same convention as
`modellib("Franken_2017_midazolam")`.

ECMO circuit blood flow rate (`Q_ECMO`, L/min) acts **only** on the metabolite's
clearance. Nothing in the parent's disposition depends on flow; this is a
structural consequence of Table 1, which reports the flow coefficient on
`CL_1-OH MDZ` alone. It has a strong bearing on the figure replications below.

```{r structure}
ui_ecmo$state
```

# Source trace

Every `ini()` value carries an in-file comment naming its source location. The
table below collects them for review. All values come from Table 1 of the main
paper, whose two column pairs give the on-ECMO and post-ECMO estimates.

| Parameter | On-ECMO | Post-ECMO | Source location |
|---|---|---|---|
| `lvc` (Vc,MDZ, L) | 36.7 | 8.25 | Table 1, Fixed effects |
| `lvp` (Vp,MDZ, L) | 57.1 | 37.4 | Table 1, Fixed effects |
| `lq` (Q,MDZ, L/h) | 3.1 | 7.88 | Table 1, Fixed effects |
| `lcl_form_1ohm` (CL_MF, L/h) | 3.26 | 5.44 | Table 1, Fixed effects |
| `lcl_1ohm` (CL_1-OH, L/h) | 49.7 | 56.6 | Table 1, Fixed effects |
| `lvc_1ohm` (Vc,1-OH, L) | 6.42 | 2.71 | Table 1, Fixed effects |
| `lvp_1ohm` (Vp,1-OH, L) | 127 | 98 | Table 1, Fixed effects |
| `lq_1ohm` (Q,1-OH, L/h) | 0.443 | 1.14 | Table 1, Fixed effects |
| `e_q_ecmo_cl_1ohm` | 0.336 | n/a | Table 1, `theta ECMO flow rate on CL1-OH MDZ` |
| centering median for `Q_ECMO` | 2.7 L/min | n/a | **Not published** -- back-solved from Fig. 1A (see Errata) |
| covariate functional form | median-normalized proportional | n/a | **Not published** -- recovered by fitting Fig. 1A (see Errata) |
| `etalcl_1ohm` | 0.246 | 0.574 | Table 1, Random effects (variance scale) |
| `etalcl_form_1ohm` | 0.497 | 0.732 | Table 1, Random effects |
| `etalvc` | 0.693 | 0 FIX (omitted) | Table 1, Random effects |
| `etalvp` | 0.176 | 1.08 | Table 1, Random effects |
| `etalvc_1ohm` | 0.137 | 0.118 | Table 1, Random effects |
| `propSd` (parent) | sqrt(0.404) | sqrt(0.525) | Table 1, Residual variability |
| `propSd_1ohm` | sqrt(0.362) | sqrt(0.359) | Table 1, Residual variability |
| ODE structure | -- | -- | Supplementary Fig. S1 |
| proportional error model | -- | -- | Supplementary Methods, "Population pharmacokinetic model development" |

## Random effects are on the variance scale

Table 1's footnote says random effects are "presented as estimates with CV %",
which is ambiguous about whether the tabulated number is a variance or a
standard deviation. It is a **variance**: the parenthetical is exactly
`100 * sqrt(omega^2)` for all nine random-effect entries across both models.

```{r variance-scale}
omega2 <- c(0.246, 0.497, 0.693, 0.176, 0.137, 0.574, 0.732, 1.08, 0.118)
printed_cv <- c(49.6, 70.5, 83.3, 42.0, 37.0, 75.8, 85.6, 103.9, 34.4)
data.frame(
  omega2 = omega2,
  implied_cv_pct = round(100 * sqrt(omega2), 1),
  printed_cv_pct = printed_cv
)
stopifnot(all(abs(100 * sqrt(omega2) - printed_cv) < 0.06))
```

All nine match to better than 0.06 percentage points, so the tabulated numbers
are used directly as variances.

# Validation strategy

The paper reports **no NCA table** -- no Cmax, no AUC, no half-life -- so there
is nothing to transcribe for a published-vs-simulated NCA comparison. Instead
this vignette validates against three things that are model-independent and
exactly checkable:

1. **Closed-form steady-state and AUC identities.** Because `CL_MF` is the only
   parent elimination and the parent-to-metabolite transfer is 1:1, mass balance
   forces `AUC_inf(parent) = Dose / CL_MF` and `AUC_inf(1-OH) = Dose / CL_1-OH`
   exactly. These are computed with PKNCA below.
2. **Figure 1A**, which carries exact numeric labels and is the paper's headline
   quantitative result.
3. **Figure 1B**, which turns out **not** to be reproducible from Table 1 -- a
   genuine internal inconsistency in the paper, documented in the Errata with
   the sensitivity analysis that isolates the cause.

# Virtual cohort

Original observed data are not publicly available. All simulations below use
virtual cohorts. Typical-value simulations use `rxode2::zeroRe()`; the
between-subject-variability cohort uses 100 subjects per ECMO-flow arm, well
under the 200-per-arm cap.

```{r helpers}
# Graded observation grid: dense through distribution, sparse in the tail.
obs_grid <- function(tmax = 720) {
  unique(sort(c(
    seq(0, 4, by = 0.05),
    seq(4, 24, by = 0.25),
    seq(24, 240, by = 1),
    seq(240, tmax, by = 4)
  )))
}

# One arm = one ECMO flow rate. `id_offset` keeps IDs disjoint across arms.
make_arm <- function(flow, n = 1L, amt, rate = NULL, times, id_offset = 0L,
                     label = paste0(flow, " L/min")) {
  ids <- id_offset + seq_len(n)
  doses <- tidyr::expand_grid(id = ids, time = 0) |>
    dplyr::mutate(amt = amt, evid = 1L, cmt = "central",
                  rate = if (is.null(rate)) 0 else rate, dvid = 1L)
  obs <- tidyr::expand_grid(id = ids, time = times) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central",
                  rate = 0, dvid = 1L)
  dplyr::bind_rows(doses, obs) |>
    dplyr::mutate(Q_ECMO = flow, arm = label) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}
```

Observation rows use `cmt = "central"` -- the ODE state, never the algebraic
observable name `Cc`. `dvid = 1L` on every observation row makes rxode2 return
both endpoint columns (`Cc` and `Cc_1ohm`).

# Replicate Figure 1A

Figure 1A reports the midazolam infusion rate needed to achieve the same 24-h
plasma concentration as a 1 mg/h infusion at 1 L/min, at ECMO flows of 1-5
L/min. The five plotted points are labelled 1.00, 1.16, 1.34, 1.47, and 1.63.

Because the system is linear in dose, the required infusion-rate ratio equals
the inverse ratio of the 24-h concentration achieved at a fixed 1 mg/h infusion.

```{r figure-1a}
flows <- 1:5

sim_24h <- lapply(seq_along(flows), function(i) {
  ev <- make_arm(
    flow = flows[i], n = 1L,
    amt = 24, rate = 1,                    # 1 mg/h for 24 h
    times = seq(0, 24, by = 0.25),
    id_offset = (i - 1L) * 10L
  )
  s <- rxode2::rxSolve(
    rxode2::zeroRe(ui_ecmo), events = ev,
    keep = c("arm", "Q_ECMO"), useLinCmt = FALSE, returnType = "data.frame"
  )
  s[which.min(abs(s$time - 24)), ]
}) |> dplyr::bind_rows()

fig1a <- sim_24h |>
  dplyr::transmute(
    flow      = Q_ECMO,
    C24_1ohm  = Cc_1ohm,
    C24_parent = Cc,
    simulated = C24_1ohm[1] / C24_1ohm,
    published = c(1.00, 1.16, 1.34, 1.47, 1.63)
  ) |>
  dplyr::mutate(difference = simulated - published)

fig1a |>
  dplyr::mutate(dplyr::across(C24_1ohm:difference, \(x) round(x, 4))) |>
  dplyr::rename(
    "ECMO flow (L/min)"          = flow,
    "24 h 1-OH conc (ng/mL)"     = C24_1ohm,
    "24 h midazolam conc (ng/mL)" = C24_parent,
    "Simulated ratio"            = simulated,
    "Published (Fig. 1A)"        = published,
    "Difference"                 = difference
  ) |>
  knitr::kable(caption = "Replicates Figure 1A of Kim 2026.")

rmse_1a <- sqrt(mean(fig1a$difference^2))
rmse_1a
stopifnot(rmse_1a < 0.02, max(abs(fig1a$difference)) < 0.03)
```

```{r figure-1a-plot}
fig1a |>
  tidyr::pivot_longer(c(simulated, published),
                      names_to = "source", values_to = "ratio") |>
  ggplot(aes(flow, ratio, colour = source, shape = source)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 1:5) +
  labs(x = "ECMO flow rate (L/min)",
       y = "Relative infusion rate (mg/h)",
       colour = NULL, shape = NULL,
       title = "Figure 1A - infusion rate needed to hold the 24 h concentration",
       caption = "Replicates Figure 1A of Kim 2026.")
```

The packaged model reproduces all five published labels with an RMSE of
`r round(rmse_1a, 4)` and a maximum absolute deviation of
`r round(max(abs(fig1a$difference)), 4)`. This is the end-to-end confirmation of
the reconstructed covariate form and centering median (see Errata): it runs
through the full ODE system rather than the algebraic ratio that was used to
recover them.

Note the fourth column of the table above: the **midazolam** concentration at
24 h is identical at every flow. Figure 1A's caption and the main text both say
"midazolam", but a midazolam-based figure would be a flat line at 1.00, because
no parameter in the parent's disposition depends on flow. Figure 1A must
therefore be a 1-OH-midazolam result. See the Errata.

# Closed-form identities

Mass balance gives two exact checks. At steady state under a constant infusion
rate `R`, the parent concentration is `R / CL_MF` and -- because every cleared
milligram of parent becomes a milligram-equivalent of metabolite -- the
metabolite concentration is `R / CL_1-OH`.

```{r steady-state}
ev_ss <- make_arm(flow = 2.7, n = 1L, amt = 600, rate = 1,
                  times = seq(0, 600, by = 2))
ss <- rxode2::rxSolve(rxode2::zeroRe(ui_ecmo), events = ev_ss,
                      useLinCmt = FALSE, returnType = "data.frame")
ss_end <- ss[ss$time == 598, ]

ss_check <- data.frame(
  analyte   = c("midazolam", "1-OH-midazolam"),
  simulated = c(ss_end$Cc, ss_end$Cc_1ohm),
  closed_form = c(1 / 3.26 * 1000, 1 / 49.7 * 1000)
) |>
  dplyr::mutate(rel_diff_pct = 100 * (simulated / closed_form - 1))

ss_check |>
  dplyr::mutate(dplyr::across(simulated:rel_diff_pct, \(x) round(x, 3))) |>
  dplyr::rename(
    "Analyte"                 = analyte,
    "Simulated Css (ng/mL)"   = simulated,
    "R / CL (ng/mL)"          = closed_form,
    "Relative difference (%)" = rel_diff_pct
  ) |>
  knitr::kable(caption = "Steady-state identity at a 1 mg/h infusion, ECMO flow 2.7 L/min.")

stopifnot(all(abs(ss_check$rel_diff_pct) < 0.2))
```

Both hold to better than 0.2%, confirming the ODE encoding, the 1:1 mass
transfer, and the mg-to-ng/mL scaling.

# PKNCA validation

NCA is run on typical-value profiles after a single 5 mg IV bolus at each ECMO
flow. Under the mass-balance argument above, `AUC_inf` must equal `Dose / CL_MF`
for the parent and `Dose / CL_1-OH` for the metabolite, so PKNCA output can be
compared against an exact reference rather than a tolerance band.

```{r simulate-nca}
dose_mg <- 5
grid <- obs_grid(720)

events_nca <- lapply(seq_along(flows), function(i) {
  make_arm(flow = flows[i], n = 1L, amt = dose_mg, times = grid,
           id_offset = (i - 1L) * 10L)
}) |> dplyr::bind_rows()

stopifnot(!anyDuplicated(events_nca[, c("id", "time", "evid")]))

# All five arms are solved in ONE rxSolve call. Solving them separately and
# binding afterwards would silently lose the `id` column: rxode2 omits `id`
# when a solve contains a single subject, so a per-arm lapply() returns
# frames with no subject identifier for PKNCA to group on.
sim_nca_raw <- rxode2::rxSolve(
  rxode2::zeroRe(ui_ecmo), events = events_nca,
  keep = c("arm", "Q_ECMO"), useLinCmt = FALSE, returnType = "data.frame"
)

stopifnot("id" %in% names(sim_nca_raw), dplyr::n_distinct(sim_nca_raw$id) == length(flows))
```

```{r pknca-parent}
conc_parent <- sim_nca_raw |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

conc_parent <- dplyr::bind_rows(
  conc_parent,
  conc_parent |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, arm, time)

dose_df <- events_nca |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, arm)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

nca_parent <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_parent, Cc ~ time | arm + id,
                   concu = "ng/mL", timeu = "h"),
  PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id, doseu = "mg"),
  intervals = intervals
))
```

```{r pknca-metabolite}
conc_metab <- sim_nca_raw |>
  dplyr::filter(!is.na(Cc_1ohm)) |>
  dplyr::select(id, time, Cc = Cc_1ohm, arm)

conc_metab <- dplyr::bind_rows(
  conc_metab,
  conc_metab |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, arm, time)

nca_metab <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_metab, Cc ~ time | arm + id,
                   concu = "ng/mL", timeu = "h"),
  PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id, doseu = "mg"),
  intervals = intervals
))
```

## Simulated NCA versus the closed-form reference

The reference column below is **not** a published NCA table -- Kim 2026 reports
none. It is the exact mass-balance identity `Dose / CL`, evaluated per arm with
the ECMO-flow effect applied to `CL_1-OH`.

```{r nca-reference}
cl_mf   <- 3.26
cl_1ohm <- 49.7 * (1 + 0.336 * (flows - 2.7) / 2.7)

ref_parent <- tibble::tibble(
  arm         = paste0(flows, " L/min"),
  aucinf.obs  = dose_mg / cl_mf * 1000
)
ref_metab <- tibble::tibble(
  arm         = paste0(flows, " L/min"),
  aucinf.obs  = dose_mg / cl_1ohm * 1000
)

cmp_parent <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_parent, reference = ref_parent, by = "arm",
  units = c(aucinf.obs = "ng*h/mL"), tolerance_pct = 1
)
cmp_metab <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_metab, reference = ref_metab, by = "arm",
  units = c(aucinf.obs = "ng*h/mL"), tolerance_pct = 1
)

knitr::kable(
  cmp_parent,
  caption = paste(
    "Midazolam AUC0-inf: PKNCA versus the closed-form Dose / CL_MF.",
    "* differs from reference by >1%."
  ),
  align = c("l", "l", "r", "r", "r")
)

knitr::kable(
  cmp_metab,
  caption = paste(
    "1-OH-midazolam AUC0-inf: PKNCA versus the closed-form Dose / CL_1-OH.",
    "* differs from reference by >1%."
  ),
  align = c("l", "l", "r", "r", "r")
)
```

```{r nca-assert}
pct_col <- function(x) {
  nm <- grep("diff", names(x), ignore.case = TRUE, value = TRUE)[1]
  as.numeric(gsub("[^0-9.eE+-]", "", as.character(x[[nm]])))
}
stopifnot(all(abs(pct_col(cmp_parent)) < 1, na.rm = TRUE))
stopifnot(all(abs(pct_col(cmp_metab)) < 1, na.rm = TRUE))
```

Both analytes match their closed-form AUC to within 1% at every flow, using a
tolerance twenty times tighter than the usual 20% NCA band. The
metabolite AUC falls monotonically with ECMO flow, reproducing the direction of
the trend in Supplementary Fig. S7.

```{r nca-summary}
dplyr::bind_rows(
  as.data.frame(nca_parent$result) |> dplyr::mutate(analyte = "midazolam"),
  as.data.frame(nca_metab$result) |> dplyr::mutate(analyte = "1-OH-midazolam")
) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "half.life")) |>
  dplyr::select(analyte, arm, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 2))) |>
  dplyr::rename(
    "Analyte"          = analyte,
    "ECMO flow"        = arm,
    "Cmax (ng/mL)"     = cmax,
    "Tmax (h)"         = tmax,
    "t1/2 (h)"         = half.life
  ) |>
  knitr::kable(caption = "Simulated NCA after a single 5 mg IV midazolam bolus.")
```

# Figure 1B does not reproduce from Table 1

Figure 1B reports the time for plasma concentrations to fall 70% after a 24-h
continuous infusion, labelled 20.62, 19.35, 17.45, 16.81, and 15.90 h at flows
of 1-5 L/min. The packaged model does not reproduce that gradient.

```{r figure-1b}
decrement_time <- function(ui, flow, column, frac = 0.30) {
  ev <- make_arm(flow = flow, n = 1L, amt = 24, rate = 1,
                 times = unique(sort(c(seq(0, 24, by = 0.5),
                                       seq(24, 120, by = 0.02)))))
  s <- rxode2::rxSolve(rxode2::zeroRe(ui), events = ev,
                       useLinCmt = FALSE, returnType = "data.frame")
  post <- s[s$time >= 24, ]
  y <- post[[column]]
  post$time[which(y <= frac * y[1])[1]] - 24
}

fig1b <- data.frame(
  flow          = flows,
  sim_1ohm      = vapply(flows, decrement_time, numeric(1),
                         ui = ui_ecmo, column = "Cc_1ohm"),
  sim_midazolam = vapply(flows, decrement_time, numeric(1),
                         ui = ui_ecmo, column = "Cc"),
  published     = c(20.62, 19.35, 17.45, 16.81, 15.90)
)

fig1b |>
  dplyr::mutate(dplyr::across(sim_1ohm:published, \(x) round(x, 2))) |>
  dplyr::rename(
    "ECMO flow (L/min)"        = flow,
    "Simulated, 1-OH (h)"      = sim_1ohm,
    "Simulated, midazolam (h)" = sim_midazolam,
    "Published (Fig. 1B, h)"   = published
  ) |>
  knitr::kable(caption = "Figure 1B of Kim 2026 versus the packaged model.")
```

The simulated 70% decrement time is essentially flat -- it moves by
`r round(diff(range(fig1b$sim_1ohm)), 2)` h across the whole flow range, against
a published swing of `r round(diff(range(fig1b$published)), 2)` h. The *level* is right:
the published values average `r round(mean(fig1b$published), 2)` h and the
simulated 1-OH values average `r round(mean(fig1b$sim_1ohm), 2)` h. Only the
slope is missing.

The cause is structural, not a covariate-encoding error. 1-OH-midazolam is
formation-rate-limited: `CL_MF` is 3.26 L/h against a parent volume of roughly
94 L, so the metabolite's post-infusion decline is governed by the parent's
disposition, which carries no flow dependence at all. Varying `CL_1-OH` alone
over a five-fold range barely moves the decrement time:

```{r figure-1b-sensitivity}
sens <- vapply(c(25, 49.7, 90, 130), function(cl) {
  ui_alt <- rxode2::ini(rxode2::zeroRe(ui_ecmo), lcl_1ohm = log(cl))
  decrement_time(ui_alt, flow = 2.7, column = "Cc_1ohm")
}, numeric(1))

data.frame(
  cl_1ohm = c(25, 49.7, 90, 130),
  decrement_h = round(sens, 2)
) |>
  dplyr::rename(
    "CL_1-OH (L/h)"          = cl_1ohm,
    "70% decrement time (h)" = decrement_h
  ) |>
  knitr::kable(caption = "Sensitivity of the 70% decrement time to CL_1-OH alone.")

stopifnot(diff(range(sens)) < 0.5)
```

A five-fold change in `CL_1-OH` moves the decrement time by under
`r round(diff(range(sens)), 2)` h. No value of the flow coefficient, and no
choice among the exponential, power, or proportional covariate forms, can
generate the published 4.72-h gradient from the Table 1 parameters. The
discrepancy is recorded in the Errata; the model is **not** tuned to match.

# Between-subject variability

Interindividual variability is simulated for three flow arms at 100 subjects
each. Each stochastic block is seeded immediately before it runs, because
`set.seed()` alone does not control rxode2's eta draws across separate solves.

```{r vpc}
vpc_flows <- c(1, 3, 5)
n_per_arm <- 100

events_vpc <- lapply(seq_along(vpc_flows), function(i) {
  make_arm(flow = vpc_flows[i], n = n_per_arm, amt = 24, rate = 1,
           times = seq(0, 72, by = 1),
           id_offset = (i - 1L) * n_per_arm)
}) |> dplyr::bind_rows()

stopifnot(!anyDuplicated(events_vpc[, c("id", "time", "evid")]))

set.seed(20260824)
sim_vpc <- rxode2::rxSolve(
  ui_ecmo, events = events_vpc, keep = c("arm", "Q_ECMO"),
  useLinCmt = FALSE, returnType = "data.frame"
)
```

```{r vpc-plot}
sim_vpc |>
  dplyr::filter(time > 0) |>
  tidyr::pivot_longer(c(Cc, Cc_1ohm), names_to = "analyte", values_to = "conc") |>
  dplyr::mutate(analyte = ifelse(analyte == "Cc", "midazolam", "1-OH-midazolam")) |>
  dplyr::group_by(analyte, arm, time) |>
  dplyr::summarise(
    Q05 = quantile(conc, 0.05), Q50 = median(conc),
    Q95 = quantile(conc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(time, Q50, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~analyte, scales = "free_y") +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Concentration (ng/mL)",
       colour = "ECMO flow", fill = "ECMO flow",
       title = "Simulated variability, 1 mg/h infusion for 24 h then stop",
       caption = "Median and 5th-95th percentiles, 100 virtual subjects per arm.")
```

The flow arms separate for 1-OH-midazolam and are superimposed for midazolam,
which is the structural signature of a covariate acting on the metabolite's
clearance alone.

# On-ECMO versus post-ECMO

The two models were fit independently, so a direct comparison of their typical
profiles is the natural summary of the paper's clinical message. The post-ECMO
model has a much smaller central volume (8.25 L vs 36.7 L) and a faster
formation clearance (5.44 L/h vs 3.26 L/h).

```{r phase-comparison}
ev_cmp <- make_arm(flow = 2.7, n = 1L, amt = 24, rate = 1,
                   times = seq(0, 72, by = 0.5))

sim_phase <- dplyr::bind_rows(
  rxode2::rxSolve(rxode2::zeroRe(ui_ecmo), events = ev_cmp,
                  useLinCmt = FALSE, returnType = "data.frame") |>
    dplyr::mutate(phase = "on ECMO (flow 2.7 L/min)"),
  rxode2::rxSolve(rxode2::zeroRe(ui_post), events = ev_cmp,
                  useLinCmt = FALSE, returnType = "data.frame") |>
    dplyr::mutate(phase = "after ECMO")
)

sim_phase |>
  dplyr::filter(time > 0) |>
  tidyr::pivot_longer(c(Cc, Cc_1ohm), names_to = "analyte", values_to = "conc") |>
  dplyr::mutate(analyte = ifelse(analyte == "Cc", "midazolam", "1-OH-midazolam")) |>
  ggplot(aes(time, conc, colour = phase)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~analyte, scales = "free_y") +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Concentration (ng/mL)", colour = NULL,
       title = "Typical profiles, 1 mg/h midazolam for 24 h then stop",
       caption = "modellib('Kim_2026_midazolam_ecmo') vs modellib('Kim_2026_midazolam_postecmo').")
```

# Assumptions and deviations

## Errata and reporting gaps in the source

* **The covariate functional form is never stated.** Supplementary Methods says
  continuous covariates were "centered on median values and evaluated using
  exponential, power, and proportional models" but never says which form was
  retained for ECMO flow rate. The **median-normalized proportional** form was
  recovered by fitting the paper's own Figure 1A. Of the candidates, only it
  fits: RMSE 0.011 against the five published labels, versus 0.094 for the power
  form; the plain proportional form `1 + theta * (F - Fm)` would require a
  negative median of -2.3 L/min, and the exponential form overshoots grossly
  (predicting 1.40 at 2 L/min against a published 1.16). Two independent
  corroborations: the main text describes the relationship as a "linear rise",
  which excludes the power and exponential forms, and the Methods' own wording
  specifies median centering.

* **The centering median is published nowhere.** Supplementary Table S1
  tabulates every other screened covariate -- age, weight, sex, indication, AST,
  ALT, serum creatinine, ECMO duration -- but omits ECMO flow rate. The value
  **2.7 L/min is figure-derived, not paper-reported**: solving for the median at
  each of the four non-trivial Figure 1A points gives 2.66, 2.47, 2.72, and 2.71
  L/min. This is flagged inline in the model file's `ini()` comment and in
  `covariateData[[Q_ECMO]]$notes`. It is a plausible adult VA-ECMO flow, and the
  Figure 1A replication above confirms it end-to-end through the ODE system, but
  a reader who obtains the control stream should verify it.

* **Figure 1B is not reproducible from Table 1.** See the dedicated section
  above. The published 70%-decrement gradient (20.62 h down to 15.90 h) cannot
  be generated by any covariate form acting on `CL_1-OH`, because the metabolite
  is formation-rate-limited and its decline tracks the flow-independent parent.
  The published across-flow *mean* (`r round(mean(fig1b$published), 2)` h) does
  match the simulated value (`r round(mean(fig1b$sim_1ohm), 2)` h). The model
  reproduces Table 1 faithfully; the inconsistency is in the source.

* **The paper contradicts itself about which analyte Figures 1A and 1B show.**
  The Figure 1B caption, the Supplementary Fig. S6 caption, and the
  Supplementary Methods all say "midazolam"; the main text Results paragraph
  says "1-hydroxymidazolam". It must be the metabolite -- the covariate acts
  only on `CL_1-OH`, so a midazolam-based version of either figure would be flat
  across flow. The Figure 1A table above demonstrates exactly this: simulated
  midazolam concentration at 24 h is identical at all five flows. Supplementary
  Fig. S8, which plots 1-OH profiles with the 70% points marked, corroborates
  the metabolite reading.

* **An ETA correlation is reported for an eta that has no reported variance.**
  Table 1 has a row under "ETA correlation" labelled
  `omega^2 CL1-OH MDZ, omega^2 Q1-OH MDZ` with a value of 0.228 (bootstrap
  95% CI 0-0.56, on-ECMO model only). The Random effects block above it reports
  variances for five etas, and **none of them is an eta on `Q_1-OH MDZ`**. The
  correlation therefore names an eta whose variance is published nowhere and
  cannot be built without inventing one, so it is **not encoded**. Note also
  that if the row were instead a covariance between the two 1-OH etas that do
  exist, 0.228 would imply a correlation of `0.228 / sqrt(0.246 * 0.137)` =
  1.24, which is impossible -- so under any reading the number is a correlation
  coefficient rather than a covariance. Its bootstrap CI includes zero. The row
  is recorded here verbatim so a reader holding the control stream can restore
  it.

* **Supplementary Fig. S7's y-axis units appear to be wrong by a factor of
  1000.** The axis reads "AUC0-48 of 1-OH MDZ (ng*h/mL)" but the plotted values
  (roughly 0.4-0.7 at a 1 mg/h infusion) are consistent with mg*h/L, i.e.
  ug*h/mL. Only the trend direction from that figure is used here.

* **Supplementary Table S2 tabulates a noisy, non-monotonic "time to target
  concentration"** (13-21 h) across the 50 dose-by-flow scenarios, which is not
  consistent with the smooth monotonic curve of Figure 1B. Those values come
  from Monte Carlo simulation with between-subject variability (500 replicates
  of 50 scenarios per Supplementary Methods), so they are not directly
  comparable to a typical-value reproduction and are not used as a validation
  target.

## Encoding decisions

* **Parent-to-metabolite transfer is 1:1 in midazolam-equivalent mass.** The
  paper states no molar conversion factor and reports no fraction metabolised,
  and `CL_MF` is the only parent elimination pathway, so all parent mass is
  transferred. This matches the convention used by
  `modellib("Franken_2017_midazolam")`. A true molar correction
  (MW 341.8 / 325.8 = 1.049) would rescale `CL_1-OH` and the 1-OH volumes by
  under 5%, and cannot be applied without knowing whether the assay already
  reported midazolam equivalents.

* **Random effects are read as variances, not standard deviations.** Justified
  in the Source trace section above: the printed parenthetical is
  `100 * sqrt(omega^2)` for all nine entries. Independently, the residual RSEs
  are consistent with a variance scale -- the RSE of an estimated variance is
  roughly `sqrt(2/N)`, and the parent's 13.3% implies about 113 observations,
  plausible for 19 serially sampled patients, whereas an SD scale would imply
  about 28.

* **The post-ECMO `omega^2 Vc,MDZ = 0 FIX` is omitted rather than written as
  `fixed(0)`.** A zero-variance diagonal makes OMEGA singular and breaks the
  Cholesky sampler that `rxSolve()` uses, and a zero-variance eta contributes
  nothing, so the two encodings are mathematically identical. Same handling as
  `modellib("Wattanakul_2024_primaquine_motherinfant")`.

* **Concentrations are reported in ng/mL**, matching the assay described in
  Supplementary Methods (LLOQ 25 ng/mL midazolam, 5 ng/mL 1-OH-midazolam). The
  ODE states hold mg and the volumes are litres, so `model()` scales
  amount/volume by 1000.

* **Evaluated-but-not-retained covariates** are recorded in each model's
  `covariatesDataExcluded` metadata rather than `covariateData`, because the
  paper reports no effect estimate for any of them. Nine come from the
  screened-covariate list in Supplementary Methods, "Covariate model
  development" (age, weight, albumin, AST, ALT, serum creatinine, sex,
  concomitant sufentanil, concomitant remifentanil). `ECMO_PUMP_SPEED` is
  listed alongside them with a narrower claim: the supplement documents pump
  speed as *collected* ("Clinical data collection" and "ECMO system") but does
  not name it among the screened covariates, and it does not appear in Table 1.
  It is worth singling out because Kim 2026 recorded both pump speed and flow
  rate and retained only flow rate, the exact reverse of
  `modellib("Yang_2017_remifentanil")`, which tested both and retained only
  pump speed.

* **`checkModelConventions()` emits one known false positive on this model.**
  It flags `e_q_ecmo_cl_1ohm` as "a reversed-order covariate effect
  (`e_<param>_<cov>`)". The name is in fact the canonical `e_<cov>_<param>`
  order, with covariate `q_ecmo` and parameter `cl_1ohm`. The heuristic
  misfires because it splits on underscores and finds that the first token,
  `q`, is a bare PK parameter name (intercompartmental clearance) while a
  later token, `ecmo`, is a known covariate token -- a collision that is
  unavoidable for any covariate whose canonical name begins with `Q_`. The
  parameter is deliberately **not** renamed, because any name that dodges the
  heuristic would no longer match the covariate's canonical spelling. The
  check reports it as a warning, not an error, so it does not block
  `buildModelDb()`.

* **`Q_ECMO` is a new canonical covariate column**, registered in
  `inst/references/covariate-columns.md` alongside this extraction and named
  into the `Q_` circuit-flow family (with `Q_CVVH`, `Q_TOTAL_LPM`, `QBL`)
  rather than the device-prefixed `ECMO_` family. It is the first time the
  library retains ECMO flow rate as a covariate in a final model.

## Simulation assumptions

* No observed data are available, so every figure uses virtual cohorts.
* The paper does not report the observed distribution of ECMO flow rates, so the
  flow arms are the 1-5 L/min integer grid the paper itself simulates.
* `Q_ECMO` is treated as constant per subject; it is naturally time-varying
  (flow is titrated during weaning) but the paper does not state the time
  resolution used in the fit.
* Dosing is simulated as a clean 1 mg/h infusion or a single 5 mg bolus. Real
  patients received physician-titrated infusions with supplementary boluses.
