---
title: "Camostat mesylate (Kim 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Camostat mesylate (Kim 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Kim_2023_camostat"))$reference`
* Description: `r rxode2::rxode(readModelDb("Kim_2023_camostat"))$description`
* Article: <https://doi.org/10.3390/pharmaceutics15092357>
* Supplement (Table S1, the only place any estimate appears):
  <https://www.mdpi.com/article/10.3390/pharmaceutics15092357/s1>

Camostat mesylate is a serine-protease (TMPRSS2) inhibitor used in Korea for
chronic pancreatitis and post-gastrectomy reflux esophagitis, and investigated
as a COVID-19 therapy. It is hydrolysed in blood by carboxylesterase far too
fast to be quantified, so the measured analytes are its active metabolite
**GBPA** (4-(4-guanidinobenzoyloxy)phenylacetic acid, FOY-251) and the
downstream inactive metabolite **GBA** (4-guanidinobenzoic acid). GBA
circulates at roughly twice the Cmax and four times the AUC of GBPA, so the
authors fitted the two analytes jointly.

# Population

Fifteen healthy Korean adult men received a single oral camostat mesylate
tablet of 100, 200 or 300 mg under fasting conditions (n = 5 per dose) in a
parallel-group, open-label Phase 1 study (ClinicalTrials.gov NCT04782505).
Plasma was sampled at 0, 0.25, 0.5, 0.75, 1, 1.5, 2, 2.5, 3, 4, 5 and 6 h after
the dose, giving 180 samples per analyte and the 360 observations used in the
joint fit. Baseline demographics (paper Table 1) were a median age of 26 years,
median weight 66.5 kg and mean height 174.2 cm; enrolment required age 19-55
years, weight 55.0-90.0 kg, BMI 18.0-29.9 kg/m^2 and eGFR at or above
60 mL/min/1.73 m^2. All subjects were male, so the model carries no information
about female pharmacokinetics, a limitation the authors state explicitly.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Kim_2023_camostat")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Kim_2023_camostat.R`. The table below
collects them in one place for review. Every estimate is from **Supplementary
Table S1**; no parameter estimate appears anywhere in the main text.

| Equation / parameter | Value | Source location |
|---|---|---|
| Model topology (depot -> GBPA central -> GBA central) | n/a | Figure 4 (p. 7) schematic and its caption |
| `lka` | 0.757 1/h | Table S1, Ka |
| `ltlag` | 0.204 h | Table S1, Tlag |
| `lvc` | 237.9 L | Table S1, Vd1 |
| `lcl` | 1269 L/h | Table S1, CL1 (unit column blank; see Errata) |
| `lvc_gba` | 174.6 L | Table S1, Vd2 |
| `lcl_gba` | 103.7 L/h | Table S1, CL2 (Figure 4 CLmet) |
| `fm` | 0.816 | Table S1, FRAC = 0.184; Figure 4 uses (1 - FRAC) on the metabolic branch; Discussion quotes "about 82%" |
| `lvmax` | 650 mg/h | **NOT from the paper.** Derived from Table 2 and Figure 6; see Errata |
| `lkm` | 1192 ng/mL | Table S1, Km = 1.192 read as mg/L; see Errata |
| `etalka` | 0.109^2 | Table S1, BSV Ka |
| `etaltlag` | 0.089^2 | Table S1, BSV Tlag |
| `etalcl` | 0.629^2 | Table S1, BSV CL1 |
| `etalcl_gba` | 0.210^2 | Table S1, BSV CL2 |
| `etalvc` + `etalvmax` block | 0.813^2, -0.89, 0.820^2 | Table S1, BSV V1, BSV Vmax, Corr_Vmax_V1 |
| `propSd` | 0.25 | Table S1, Pro_GBPA |
| `propSd_gba` | 0.23 | Table S1, Pro_GBA |
| NCA reference values (Cmax, AUClast, AUCinf, half-life) | Table 2 | Paper Table 2 (p. 7) |
| NCA reference Tmax (median) | 1 h GBPA, 2 h GBA | Paper Section 3.2 body text |
| Multiple-dose simulation reference | Figure 6 | Paper Figure 6 (p. 8), 200 mg TID for 14 days |

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the trial design (three parallel dose arms) with 150 virtual subjects per arm.
The model carries no covariates, so the only per-subject variation is the
random-effect draw.

```{r cohort}
# set.seed() seeds R's RNG. It does NOT seed rxode2's simulation RNG, and
# rxode2's streams are partitioned per solver thread, so this cohort is
# reproducible on one machine and different on a machine with a different
# thread count. Every assertion below is written to hold for any cohort the
# model can produce.
set.seed(20230921)

n_per_arm <- 150L

# The paper's nominal sampling schedule (Section 2.3).
nominal_times <- c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 2.5, 3, 4, 5, 6)

# Dense grid for the figures, with the nominal times forced in exactly.
# Rounding avoids the floating-point drift that would otherwise make
# `time %in% nominal_times` silently match nothing later on.
obs_times <- sort(unique(round(c(nominal_times, seq(0, 16, by = 0.05)), 6)))

make_arm <- function(dose, id_offset) {
  ids <- id_offset + seq_len(n_per_arm)
  lbl <- paste0(dose, " mg")
  dplyr::bind_rows(
    data.frame(id = ids, time = 0, amt = dose, cmt = "depot",
               evid = 1L, dvid = NA_integer_, treatment = lbl),
    expand.grid(id = ids, time = obs_times) |>
      dplyr::mutate(amt = NA_real_, cmt = "central", evid = 0L,
                    dvid = 1L, treatment = lbl)
  )
}

events <- dplyr::bind_rows(
  make_arm(100, 0L),
  make_arm(200, n_per_arm),
  make_arm(300, 2L * n_per_arm)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
stopifnot(length(unique(events$id)) == 3L * n_per_arm)
```

The event table uses `cmt = "central"` on observation rows with an explicit
`dvid`, rather than naming the observable. This model has two `~` residual
endpoints, so every observation row must say which endpoint it belongs to;
`rxSolve()` returns both `Cc` and `Cc_gba` as columns regardless.

# Simulation

```{r simulate}
mod <- readModelDb("Kim_2023_camostat")

sim <- rxode2::rxSolve(
  mod, events = events,
  keep = "treatment",
  # rxode2's automatic ODE -> linCmt conversion corrupts the dvid mapping for
  # multi-output models; see known-vignette-failure-patterns pattern 5b.
  useLinCmt = FALSE
) |>
  as.data.frame()

stopifnot(all(sim$Cc >= 0), all(sim$Cc_gba >= 0))
```

A typical-value (no random effects) solve is used for the deterministic checks
and for the internal mass-balance identity.

```{r simulate-typical}
mod_typical <- rxode2::zeroRe(mod)

events_typ <- events |>
  dplyr::filter(id %in% c(1L, n_per_arm + 1L, 2L * n_per_arm + 1L))

sim_typ <- rxode2::rxSolve(
  mod_typical, events = events_typ,
  keep = "treatment",
  # Explicit no-random-effects sentinel: rxSolve() otherwise reuses the omega
  # from the stochastic solve above and silently re-samples etas.
  omega = NA, sigma = NA,
  useLinCmt = FALSE
) |>
  as.data.frame()
```

# Replicate published figures

## Figure 2 -- plasma concentration-time profiles by dose

```{r figure-2, fig.height = 5}
# Replicates Figure 2 of Kim 2023: mean (SD) GBPA and GBA plasma
# concentrations over the 0-6 h sampling window, one series per dose group.
fig2 <- sim |>
  dplyr::filter(time <= 6) |>
  dplyr::select(id, time, treatment, GBPA = Cc, GBA = Cc_gba) |>
  tidyr::pivot_longer(c(GBPA, GBA), names_to = "analyte", values_to = "conc") |>
  dplyr::group_by(treatment, analyte, time) |>
  dplyr::summarise(mean = mean(conc), sd = stats::sd(conc), .groups = "drop") |>
  dplyr::mutate(analyte = factor(analyte, levels = c("GBPA", "GBA")))

ggplot(fig2, aes(time, mean, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = pmax(mean - sd, 0), ymax = mean + sd),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~analyte) +
  labs(x = "Time (h)", y = "Plasma concentration (ng/mL)",
       colour = "Dose", fill = "Dose",
       title = "Figure 2 -- mean (SD) concentration-time profiles",
       caption = "Replicates Figure 2 of Kim 2023.") +
  theme(legend.position = "bottom")
```

## Figure 5 -- visual predictive check

Figure 5 of the paper pools all three dose groups (all 15 subjects), which is
what produces its wide spread; the panels below are constructed the same way.
Note that `Cc` and `Cc_gba` from `rxSolve()` are individual predictions and do
**not** carry the proportional residual error, so these bands are narrower than
the paper's prediction intervals by the residual component.

```{r figure-5, fig.height = 5}
# Replicates Figure 5 of Kim 2023: 5th, 50th and 95th percentiles of the
# predicted concentrations, all dose groups pooled, 0-6 h.
fig5 <- sim |>
  dplyr::filter(time <= 6) |>
  dplyr::select(time, GBPA = Cc, GBA = Cc_gba) |>
  tidyr::pivot_longer(c(GBPA, GBA), names_to = "analyte", values_to = "conc") |>
  dplyr::group_by(analyte, time) |>
  dplyr::summarise(
    Q05 = stats::quantile(conc, 0.05),
    Q50 = stats::quantile(conc, 0.50),
    Q95 = stats::quantile(conc, 0.95),
    .groups = "drop"
  ) |>
  dplyr::mutate(analyte = factor(analyte, levels = c("GBPA", "GBA")))

ggplot(fig5, aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~analyte) +
  labs(x = "Time (h)", y = "Plasma concentration (ng/mL)",
       title = "Figure 5 -- VPC, dose groups pooled",
       caption = "Replicates Figure 5 of Kim 2023.")
```

## Figure 6 -- 200 mg three times daily for 14 days

```{r figure-6, fig.height = 4}
# Replicates Figure 6 of Kim 2023: cohort-mean profiles for the first five and
# the last five of 42 doses of camostat 200 mg q8h. The paper averaged 1000
# individuals; 150 is used here (the per-arm cohort cap) and the mean is
# already stable at that size.
md_times <- c(seq(0, 40, by = 0.1), seq(304, 344, by = 0.1))
md_ids <- seq_len(n_per_arm)

events_md <- dplyr::bind_rows(
  expand.grid(id = md_ids, time = seq(0, 336, by = 8)) |>
    dplyr::mutate(amt = 200, cmt = "depot", evid = 1L, dvid = NA_integer_),
  expand.grid(id = md_ids, time = md_times) |>
    dplyr::mutate(amt = NA_real_, cmt = "central", evid = 0L, dvid = 1L)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim_md <- rxode2::rxSolve(mod, events = events_md, useLinCmt = FALSE) |>
  as.data.frame()

fig6 <- sim_md |>
  dplyr::select(time, GBPA = Cc, GBA = Cc_gba) |>
  tidyr::pivot_longer(c(GBPA, GBA), names_to = "analyte", values_to = "conc") |>
  dplyr::group_by(analyte, time) |>
  dplyr::summarise(mean = mean(conc), .groups = "drop") |>
  dplyr::mutate(
    panel = ifelse(time <= 40, "(A) first five doses", "(B) last five doses"),
    analyte = factor(analyte, levels = c("GBPA", "GBA"))
  )

ggplot(fig6, aes(time, mean, colour = analyte)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~panel, scales = "free_x") +
  scale_colour_manual(values = c(GBPA = "#2A9D8F", GBA = "#E76F51")) +
  labs(x = "Time (h)", y = "Plasma concentration (ng/mL)", colour = NULL,
       title = "Figure 6 -- camostat 200 mg TID for 14 days",
       caption = "Replicates Figure 6 of Kim 2023.") +
  theme(legend.position = "bottom")
```

```{r figure-6-peaks}
f6_peak <- fig6 |>
  dplyr::group_by(panel, analyte) |>
  dplyr::summarise(peak = max(mean), .groups = "drop")

knitr::kable(
  f6_peak |>
    dplyr::mutate(
      published = ifelse(analyte == "GBPA", 125, 283),
      pct_diff = 100 * (peak - published) / published
    ) |>
    dplyr::rename(
      "Panel" = panel, "Analyte" = analyte,
      "Simulated peak (ng/mL)" = peak,
      "Figure 6 peak (ng/mL)" = published,
      "% difference" = pct_diff
    ),
  digits = 1,
  caption = "Cohort-mean peak concentrations against the peaks read off Figure 6."
)

# The accumulation index the paper reports is 1.12 (GBPA) and 1.08 (GBA).
accum <- fig6 |>
  dplyr::group_by(analyte, panel) |>
  dplyr::summarise(peak = max(mean), .groups = "drop") |>
  tidyr::pivot_wider(names_from = panel, values_from = peak) |>
  dplyr::mutate(accumulation_index = `(B) last five doses` / `(A) first five doses`)

knitr::kable(
  accum |>
    dplyr::select(analyte, accumulation_index) |>
    dplyr::mutate(published = ifelse(analyte == "GBPA", 1.12, 1.08)) |>
    dplyr::rename("Analyte" = analyte,
                  "Peak ratio, last / first dose" = accumulation_index,
                  "Published accumulation index" = published),
  digits = 3,
  caption = paste(
    "Peak-ratio accumulation. The paper's index comes from",
    "1 / (1 - exp(-lambda_z * tau)) on the simulated profile, not from a",
    "peak ratio, so the two are related but not identical quantities."
  )
)

# Both analytes accumulate very little at an 8 h interval: the model half-lives
# are about 0.9 h (GBPA, absorption-rate limited) and 1.2 h (GBA). Observed
# range of the peak ratio across draws was 1.00-1.01; the bound below is wide
# enough to survive a different cohort and still fails if the model ever
# develops real accumulation.
stopifnot(all(accum$accumulation_index > 0.95),
          all(accum$accumulation_index < 1.30))
```

# PKNCA validation

The NCA below is run on the paper's own nominal sampling schedule (0-6 h,
12 time points) so that it reproduces the authors' procedure -- including the
AUCinf extrapolation from a 6 h window -- rather than an idealised dense-grid
NCA. One PKNCA block is run per analyte.

```{r pknca-setup}
dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "mg")

intervals <- data.frame(
  start       = 0,
  end         = Inf,
  cmax        = TRUE,
  tmax        = TRUE,
  auclast     = TRUE,
  aucinf.obs  = TRUE,
  half.life   = TRUE
)

run_nca <- function(conc_col) {
  d <- sim |>
    dplyr::filter(round(time, 6) %in% round(nominal_times, 6)) |>
    dplyr::transmute(id, time, treatment, Cc = .data[[conc_col]]) |>
    dplyr::filter(!is.na(Cc))
  # Guard the %in% filter: a floating-point mismatch would silently leave an
  # empty frame and PKNCA would then be asked to analyse nothing.
  stopifnot(nrow(d) == length(nominal_times) * 3L * n_per_arm)
  # Guarantee a time = 0 row per subject; pre-dose Cc = 0 is correct for an
  # extravascular dose.
  d <- dplyr::bind_rows(
    d,
    d |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
  ) |>
    dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
    dplyr::arrange(id, treatment, time)

  conc_obj <- PKNCA::PKNCAconc(d, Cc ~ time | treatment + id,
                               concu = "ng/mL", timeu = "h")
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}
```

```{r pknca-gbpa}
nca_gbpa <- run_nca("Cc")
nca_gba <- run_nca("Cc_gba")
```

## Comparison against published NCA

Table 2 of the paper reports arithmetic means (SD) over the five subjects in
each dose group, so the simulated results are aggregated with the arithmetic
mean rather than the median that `ncaComparisonTable()` uses by default. Tmax
is the exception: the paper reports it as a median with a range (Section 3.2,
1 h for GBPA and 2 h for GBA), so Tmax is aggregated with the median.

```{r nca-compare}
mean_by_group <- function(res) {
  d <- as.data.frame(res$result)
  if ("exclude" %in% names(d)) d <- d[is.na(d$exclude), , drop = FALSE]
  d |>
    dplyr::filter(!is.na(PPORRES)) |>
    dplyr::group_by(treatment, PPTESTCD) |>
    dplyr::summarise(
      PPORRES = if (PPTESTCD[1] == "tmax") stats::median(PPORRES) else mean(PPORRES),
      .groups = "drop"
    )
}

published_gbpa <- tibble::tribble(
  ~treatment, ~cmax,  ~tmax, ~auclast, ~aucinf.obs, ~half.life,
  "100 mg",   72.68,  1,     152.3,    156.5,       1.012,
  "200 mg",   156.8,  1,     307.4,    316.9,       1.034,
  "300 mg",   273.9,  1,     464.8,    477.5,       1.002
)

published_gba <- tibble::tribble(
  ~treatment, ~cmax,  ~tmax, ~auclast, ~aucinf.obs, ~half.life,
  "100 mg",   175.5,  2,     618.2,    762.8,       1.942,
  "200 mg",   302.9,  2,     1036,     1268,        1.954,
  "300 mg",   376.9,  2,     1270,     1710,        2.407
)

nca_units <- c(cmax = "ng/mL", tmax = "h", auclast = "h*ng/mL",
               aucinf.obs = "h*ng/mL", half.life = "h")

cmp_gbpa <- nlmixr2lib::ncaComparisonTable(
  simulated = mean_by_group(nca_gbpa),
  reference = published_gbpa,
  by = "treatment", units = nca_units, tolerance_pct = 20
)

cmp_gba <- nlmixr2lib::ncaComparisonTable(
  simulated = mean_by_group(nca_gba),
  reference = published_gba,
  by = "treatment", units = nca_units, tolerance_pct = 20
)

knitr::kable(
  cmp_gbpa,
  caption = paste(
    "GBPA: simulated vs. published NCA (Kim 2023 Table 2).",
    "* differs from the reference by more than 20%."
  )
)

knitr::kable(
  cmp_gba,
  caption = paste(
    "GBA: simulated vs. published NCA (Kim 2023 Table 2).",
    "* differs from the reference by more than 20%."
  )
)
```

**The AUC rows are the load-bearing agreement**, and they are close across
every dose group: GBPA AUC0-inf within 3.5% at all three doses, GBA AUC0-inf
within 15.5%. That is the check the derived `Vmax` has to pass (see the Errata
below), and it passes without the Figure 6 peaks -- which were not fitted --
being off by more than about 11%.

Several **Cmax, Tmax and half-life** rows do star at the 20% tolerance. None is
tuned away; each has a mechanical explanation, and the assertion block below
deliberately gates only on the AUC rows for that reason. The exact set of
starred rows shifts a little with the cohort draw, but the pattern is stable:

* **GBPA Cmax runs systematically low** (about -19 to -32% across the three
  doses). Cmax is the metric most sensitive to `Vd1`, whose between-subject SD
  is 0.813 on the log scale -- a CV of roughly 95% -- so the published
  five-subject arithmetic means are a very noisy target, and the AUC agreement
  above shows the *exposure* is right even where the peak is not.
* **GBA half-life at 300 mg.** The model's terminal rate is
  `CL2 / Vd2 = 0.594 1/h` (t1/2 = 1.17 h) and the fitted terminal slope over
  the 0-6 h window comes out near 1.84 h, whereas Table 2 reports 1.94-2.41 h
  from a five-subject regression. The model cannot produce the published 2.41 h
  from the published `CL2` and `Vd2`; this is a property of the transcribed
  parameters, not of the transcription.
* **Tmax rows star by one grid step.** Tmax is discrete on the paper's own
  12-point nominal schedule, so a single step (0.75 vs 1 h for GBPA at 100 mg,
  2.5 vs 2 h for GBA at 300 mg) is already a 25% difference. The medians
  otherwise reproduce the published 1 h (GBPA) and 2 h (GBA) exactly.

```{r nca-assertions}
# The comparison tables above hold pre-formatted strings, so the gate below
# recomputes the percent differences from the raw numbers.
pct_diff <- function(res, published, analyte) {
  published |>
    tidyr::pivot_longer(-treatment, names_to = "PPTESTCD",
                        values_to = "reference") |>
    dplyr::inner_join(mean_by_group(res), by = c("treatment", "PPTESTCD")) |>
    dplyr::transmute(analyte = analyte, treatment, PPTESTCD,
                     pct = 100 * (PPORRES - reference) / reference)
}

nca_diff <- dplyr::bind_rows(
  pct_diff(nca_gbpa, published_gbpa, "GBPA"),
  pct_diff(nca_gba, published_gba, "GBA")
)

auc_check <- nca_diff |>
  dplyr::filter(PPTESTCD %in% c("auclast", "aucinf.obs"))

# The gate must have had rows to test.
stopifnot(nrow(auc_check) == 12L)

# Structural gate: a mis-transcribed clearance, dose or unit would move the
# whole exposure distribution by tens of percent. Observed |% difference| on
# the AUC rows ran to about 20% at worst across draws; the bounds below sit
# outside that range and still go red on any real transcription error.
stopifnot(all(abs(auc_check$pct) < 35))
stopifnot(stats::median(abs(auc_check$pct)) < 20)

knitr::kable(
  nca_diff |>
    tidyr::pivot_wider(names_from = PPTESTCD, values_from = pct) |>
    dplyr::rename("Analyte" = analyte, "Dose group" = treatment),
  digits = 1,
  caption = "Percent difference from Kim 2023 Table 2, by analyte and dose group."
)
```

# Internal consistency checks

These checks use the typical-value solve, so both sides are computed from the
same parameters and the only error is numerical. Tight bounds are appropriate.

```{r mass-balance}
# Mass balance: at t = Inf every milligram of the dose has left the system
# either through the linear GBPA branch or, having become GBA, through CLmet.
#   dose = (1 - fm) * CL * AUC_GBPA  +  CLmet * AUC_GBA        (all in mg/L)
p <- rxode2::rxode(readModelDb("Kim_2023_camostat"))$theta
fm_val <- unname(p[["fm"]])
cl_val <- exp(unname(p[["lcl"]]))
cl_gba_val <- exp(unname(p[["lcl_gba"]]))

trapz <- function(x, y) sum(diff(x) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)

mb <- sim_typ |>
  # `treatment` can come back from rxSolve as a factor; as.character() keeps
  # sub() from operating on the integer level codes.
  dplyr::mutate(dose = as.numeric(sub(" mg", "", as.character(treatment)))) |>
  dplyr::group_by(treatment, dose) |>
  dplyr::summarise(
    # ng/mL -> mg/L is a factor of 1/1000.
    auc_gbpa = trapz(time, Cc) / 1000,
    auc_gba = trapz(time, Cc_gba) / 1000,
    .groups = "drop"
  ) |>
  dplyr::arrange(dose) |>
  dplyr::mutate(
    eliminated = (1 - fm_val) * cl_val * auc_gbpa + cl_gba_val * auc_gba,
    recovery = eliminated / dose
  )

knitr::kable(
  mb |>
    dplyr::select(treatment, dose, auc_gbpa, auc_gba, eliminated, recovery) |>
    dplyr::rename(
      "Dose group" = treatment, "Dose (mg)" = dose,
      "AUC GBPA (h*mg/L)" = auc_gbpa, "AUC GBA (h*mg/L)" = auc_gba,
      "Mass eliminated (mg)" = eliminated, "Recovery (fraction of dose)" = recovery
    ),
  digits = 4,
  caption = "Mass balance over the 0-16 h solve window (typical-value solve)."
)

# 16 h is over 13 GBA half-lives, so the unrecovered remainder is numerical.
stopifnot(all(abs(mb$recovery - 1) < 0.01))
```

```{r saturation}
# The Km unit assignment (1.192 mg/L = 1192 ng/mL) is what keeps the paper's
# own dose-proportionality result achievable. Confirm that the metabolic
# enzyme is only mildly saturated at the highest dose: a Km read in ng/mL
# instead would put this fraction above 0.9.
km_val <- exp(unname(p[["lkm"]]))
sat <- sim_typ |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(max_saturation = max(Cc / (km_val + Cc)), .groups = "drop")

knitr::kable(
  sat |> dplyr::rename("Dose group" = treatment,
                       "Peak Cc / (Km + Cc)" = max_saturation),
  digits = 3,
  caption = "Peak fractional saturation of the GBPA-to-GBA conversion."
)

stopifnot(all(sat$max_saturation > 0.01), all(sat$max_saturation < 0.30))

# Dose-normalised exposure moves in the direction the paper's own Table 2
# shows: GBPA AUC per 100 mg rises with dose (mild saturation of the metabolic
# branch), GBA AUC per 100 mg falls.
dn <- mb |> dplyr::mutate(gbpa_dn = auc_gbpa / dose, gba_dn = auc_gba / dose)
stopifnot(dn$gbpa_dn[3] > dn$gbpa_dn[1], dn$gba_dn[3] < dn$gba_dn[1])
```

# Assumptions and deviations

## Vmax is not the published value (non-paper provenance)

**This is the single most important caveat in this extraction.** Table S1
prints `Vmax = 2.215` (RSE 23%, bootstrap 1.97, 5-95% percentile 1.334-2.723)
and leaves the unit column blank. That row cannot be encoded as printed:

* Simulating `Vmax = 2.215 mg/h` gives a GBA AUCinf at 100 mg of
  5.8 h*ng/mL against the 762.8 h*ng/mL in the authors' own Table 2 -- under
  1%, i.e. the metabolite arm is effectively dead. It also contradicts the
  Discussion's statement that "about 82% of the GBPA metabolized to GBA in our
  model", which the printed values would put below 0.5%.
* No unit conversion closes the gap: neither x60 (1/min to 1/h) nor x1000
  (g to mg, or mg to ug) lands anywhere near the published exposures.
* An independent algebraic check that needs no simulation confirms it. Since
  `AUC = input / CL` exactly, regardless of `ka`, the published GBPA and GBA
  AUCinf values alone solve for an effective metabolic clearance of about
  500 L/h -- against the roughly 1.5 L/h implied by the printed
  `Vmax / Km`.

Exactly three rows of Table S1 have a blank unit column: `CL1`, `Vmax` and
`Km`. Every other row carries its unit. The blank cells were confirmed against
a 400 dpi render of the table, so no exponent was lost in text extraction.

The operator ruled on 2026-09-05 that `Vmax` should be **derived from the
authors' own published output and flagged as non-paper provenance**, as a
narrow exception to the standing printed-values-have-authority rule: an
unlabelled number that no unit assignment can reconcile is not a value that can
be encoded at all.

**Derivation.** With every other parameter held at its printed value,
`Vmax` was chosen to minimise the sum of squared log-ratios between this model
and the paper's six Table 2 AUCinf values (GBPA and GBA at 100, 200 and
300 mg). The objective is flat between roughly 620 and 680 mg/h; **650 mg/h**
is the minimiser and is the value shipped. It also reproduces the Figure 6
steady-state peaks, which were not fitted. The residual deviations are -4.5%,
-3.4% and -1.5% on GBPA AUCinf and -17.7%, -2.3% and +7.3% on GBA AUCinf.

Two consequences a downstream user must know:

1. `lvmax` is wrapped in `fixed()` in `ini()`. It is a derived structural
   input, not an estimate, and it must not be reported with the paper's 23%
   RSE or its bootstrap interval.
2. The effective fraction of GBPA metabolised in this encoding is about 0.66,
   not the 0.82 the Discussion quotes. The 0.82 figure is the paper's `1 -
   FRAC`, which would equal the true metabolised fraction only if
   `Vmax / Km` were exactly equal to `CL1` -- and no unit assignment makes
   that true either.

## Km unit assignment

`Km = 1.192` is likewise printed without a unit and is read here as **mg/L**
(= 1192 ng/mL). This is the only admissible reading. Observed GBPA
concentrations are 72-274 ng/mL (Table 2 Cmax), so `Km = 1192 ng/mL` leaves
the enzyme only 5-15% saturated at the peak, consistent with the
dose-proportionality the paper demonstrates (log-transformed power-model
slopes of 1.0038-1.0066 with 95% CIs inside 0.8-1.25 over 100-300 mg).
Reading `Km` as 1.192 ng/mL would leave the enzyme deeply saturated and the
model grossly dose-nonlinear. The `saturation` chunk above is the gate on this
choice. Supporting evidence: the mild saturation this reading produces has the
same sign as the paper's own dose-normalised exposures, where GBPA AUCinf per
100 mg rises across doses (156.5, 158.5, 159.2) while GBA AUCinf per 100 mg
falls (762.8, 634, 570).

## CL1 unit assignment

`CL1 = 1269` also has a blank unit cell and is retained as printed, read as
L/h. Unlike `Vmax`, that reading is dimensionally admissible and physically
plausible: Figure 4 multiplies it by a plasma concentration, and
`FRAC * CL1 = 233 L/h` is a reasonable non-metabolic apparent clearance beside
the 620-640 L/h total apparent clearance implied by Table 2's own
`Dose / AUCinf`.

## Erratum -- Table 2's CL/F row is transposed

Table 2 lists CL/F as 141.7 / 158.8 / 179.6 L/h under the GBPA columns and
718.6 / 675.1 / 680.4 L/h under the GBA columns. The body text (Section 3.2)
states the opposite: "The apparent clearance (CL/F) values for GBPA and GBA
were 704.4 L/h ... and 152.1 L/h". Computing `Dose / AUCinf` from Table 2's own
AUC row gives 639 / 631 / 628 L/h for GBPA and 131 / 158 / 175 L/h for GBA --
so the body text and the table's own arithmetic agree with each other against
the table's column labels. The `Vd` row is **not** affected
(`Vd = CL * t1/2 / ln 2` reproduces Table 2 as labelled). The CL/F row is
therefore excluded from the comparison table above; every other Table 2 row is
used as printed.

## Erratum -- BSV text/table conflict

Paper Section 3.4 says between-subject variability was applied to "lag time,
Vmax, Km, GBPA central volume of distribution (V1), clearance of GBPA, and
clearance of GBA" -- omitting `Ka` and including `Km`. Supplementary Table S1
lists a BSV row for `Ka` and has no `Km` row. The table carries the numbers
and is followed here.

## Other assumptions

* **BSV values are standard deviations, not variances.** Table S1 gives no
  scale for its between-subject variability column. Monolix (the estimation
  tool named in Section 2.6) reports `omega` as the standard deviation of the
  random effect, so the `ini()` entries are the squares of the tabulated
  values.
* **Mass equivalents.** Doses are camostat mesylate in mg and Figure 4 draws no
  molecular-weight ratio on either metabolic transfer, so every state is an
  amount in camostat mass equivalents and each volume and clearance is an
  apparent value absorbing bioavailability and the molecular-weight ratios.
  Concentrations still come out on the measured assay scale because the
  volumes and clearances were fitted against measured concentrations.
* **No covariates.** Section 3.4 reports no retained covariate effects. The
  screened-but-rejected demographics and laboratory values are recorded in the
  model file's `covariatesDataExcluded` metadata rather than in
  `covariateData`.
* **Cohort size.** The paper simulated 1000 individuals for Figure 6; 150 are
  used here, which is enough for the cohort mean to be stable and keeps the
  vignette inside its render-time budget.
* **All-male population.** The trial enrolled only men, so nothing in this
  model informs female pharmacokinetics.
