---
title: "SCR430 time-dependent pharmacokinetics in rats (Kim 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SCR430 time-dependent pharmacokinetics in rats (Kim 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Kim_2022_SCR430_rat_control_pbpk"))$reference`
* Article: <https://doi.org/10.3390/pharmaceutics14122562> (open access)

SCR430 is an investigational sorafenib derivative. In rats, 14 days of oral
SCR430 at 30 mg/kg/day lowered the exposure to a later 3 mg/kg intravenous dose
about 4.4-fold. Kim and Lee isolated hepatocytes from untreated and pretreated
rats. They found that hepatic uptake, not metabolism, had increased. They then
built a minimal PBPK (mPBPK) model in Berkeley Madonna. The model explains the
plasma and liver profiles of both groups by swapping in each group's
hepatocyte-derived inputs. The model has one structure and two parameter sets,
packaged as two files:

* `Kim_2022_SCR430_rat_control_pbpk`:
  `r rxode2::rxode(readModelDb("Kim_2022_SCR430_rat_control_pbpk"))$description`
* `Kim_2022_SCR430_rat_repeatedDosing_pbpk`:
  `r rxode2::rxode(readModelDb("Kim_2022_SCR430_rat_repeatedDosing_pbpk"))$description`

The structure is a systemic blood compartment, a "single-adjustment"
compartment exchanging with it by first-order rate constants (`k12` = paper
`kin`, `k21` = paper `kout`), and a liver of five extracellular units
(`is_liver1`..`is_liver5`) in series along hepatic blood flow. Each unit has a
hepatocellular unit (`int_liver1`..`int_liver5`) attached (paper Figure 7,
Appendix A.2 Eqs. A1 to A3). Unbound blood drug enters hepatocytes by active
uptake plus passive diffusion. Unbound hepatocellular drug leaves by passive
efflux and intrinsic metabolism.

# Population

Female Sprague-Dawley rats (8 weeks old, 190 to 205 g; Dahan Bio Link, Republic
of Korea) were split into three groups of six (paper Figure 1, Section 2.2):
untreated control, vehicle only for 14 days, and SCR430 30 mg/kg/day orally for
14 days (repeated dosing). After a 2-day washout each rat received SCR430
3 mg/kg intravenously; plasma was sampled to 10 h, and the liver was excised
at 10 h. The vehicle group did not differ from the control group (Table 1), and
the mPBPK model was built for the control and repeated-dosing groups only. The
hepatic inputs come from hepatocytes isolated from rats in each group.
The model is not a population fit, and only the systemic volume and the two
single-adjustment rate constants were fitted to plasma data.

```{r population}
str(readModelDb("Kim_2022_SCR430_rat_control_pbpk")()$population)
```

# Source trace

Every `ini()` value carries an in-file source comment; the table collects
them. "Median" rows are derived from the Table 3 (mean, SD) pairs by the
paper's own Eq. 17 (see the Assumptions section).

| Parameter | Control | Repeated dosing | Source |
|---|---|---|---|
| `vc` systemic blood volume (L/kg) | 0.018 | 0.018 | Table 3 'Central', fitted ('Equal to Control') |
| `k12` = kin (1/h) | 2.7 | 2.7 | Table 3, fitted |
| `k21` = kout (1/h) | 1.0 | 1.0 | Table 3, fitted |
| `q_liver` Qh (L/h/kg) | 3.69 | 3.69 | Table 3; Section 2.7 |
| `v_is_liver` extracellular space (L/kg) | 0.01 | 0.01 | Table 3 |
| `v_int_liver` hepatocytes (L/kg) | 0.027 | 0.027 | Table 3 (sums with the above to Liver 0.037) |
| `fup` | 0.0078 | 0.0081 | Table 1 fu,p |
| `bpr` Rb | 3.95 | 4.53 | Table 1 Rb |
| PSinf,act (L/h/kg), mean (SD) | 2.5 (0.8) | 9.2 (0.6) | Table 3 |
| PSdiff (L/h/kg), mean (SD) | 0.3 (0.09) | 0.86 (0.4) | Table 3 |
| CLint,met (L/h/kg), mean (SD) | 9.8 (1.0) | 12.3 (0.08) | Table 3 |
| fu,hepa, mean (SD) | 0.004 (0.002) | 0.01 (0.008) | Table 3; Table 1 |
| `sf_ps` scaling factor on PSinf,act and PSdiff | 1.6 | 1.6 | Section 3.7 |
| `l<param>` typical value | ln(mean) - 0.5 ln(1 + CV^2) | same | Section 2.11 Eq. 17 |
| `etal<param>` variance | ln(1 + CV^2) | same | Section 2.11 Eq. 18 |
| Extracellular units `d/dt(is_liver<i>)` | | | Appendix A.2 Eqs. A1, A2 |
| Hepatocellular units `d/dt(int_liver<i>)` | | | Appendix A.2 Eq. A3 |
| Systemic / single-adjustment `d/dt(central)`, `d/dt(peripheral1)` | | | Figure 7 schematic |
| `Cc` = systemic blood / Rb | | | Sections 2.9, 3.1 (see Assumptions) |
| `Cliver` = mean hepatocellular concentration | | | Figure 8c/d (see Assumptions) |

```{r load}
mod_ctl <- readModelDb("Kim_2022_SCR430_rat_control_pbpk")
mod_rep <- readModelDb("Kim_2022_SCR430_rat_repeatedDosing_pbpk")
ini_ctl <- rxode2::rxode(mod_ctl)$iniDf
ini_rep <- rxode2::rxode(mod_rep)$iniDf
est <- function(ini, nm) ini$est[ini$name == nm]
```

# IVIVE arithmetic (Table 2, Method 2)

The paper scales the hepatocyte uptake clearance (uL/min per 10^6 cells) to the
whole animal with 108 x 10^6 cells/g liver and 36 g liver/kg (Section 2.5).
Table 3 splits the resulting in vivo PSinf into its active and passive parts,
before the 1.6-fold factor is applied:

```{r ivive}
psf <- 108 * 36 # 10^6 cells per kg body weight
ivive <- tibble(
  group = c("Control", "Repeated dosing"),
  ps_inf_hep = c(11.9, 43.2), # Table 2, uL/min per 10^6 cells
  table2_ps_inf = c(46.2, 168.0), # Table 2, mL/min/kg
  table3_sum = c(2.5 + 0.3, 9.2 + 0.86) # Table 3 PSinf,act + PSdiff, L/h/kg
) |>
  mutate(
    scaled_ps_inf = ps_inf_hep * psf / 1000, # mL/min/kg
    table3_sum_mL_min_kg = table3_sum * 1000 / 60
  )
knitr::kable(ivive, digits = 1)
stopifnot(
  all(abs(ivive$scaled_ps_inf / ivive$table2_ps_inf - 1) < 0.01),
  all(abs(ivive$table3_sum_mL_min_kg / ivive$table2_ps_inf - 1) < 0.02)
)
```

The Table 3 sums reproduce the unscaled Method 2 PSinf in both groups. That is
why the model applies `sf_ps = 1.6` on top of the Table 3 values rather than
treating the factor as already folded in.

# Typical-value simulation and clearance check

The tandem liver is linear, so its plasma clearance has a closed form. Each of
the five units extracts `E = u / (Qh + u)` of the blood passing through,
where `u` is the unit's share of the unbound-blood intrinsic clearance,
`fub * CLint,all / 5`, and `CLint,all = PSupt * CLint,met / (PSeff + CLint,met)`
(paper Eq. 6). Blood clearance is `Qh * (1 - (1 - E)^5)`; plasma clearance is
that times Rb. The integrated model must give the same number.

```{r typical}
grid <- sort(unique(c(seq(0, 10, by = 0.05), seq(10, 96, by = 0.5))))
make_events <- function(group, n, id_offset = 0L) {
  rxode2::et(amt = 3, cmt = "central") |>
    rxode2::et(grid, cmt = "central") |>
    as.data.frame() |>
    # Two `~` endpoints (Cc, Cliver): observation rows carry dvid = 1; the
    # forward solve still returns both concentration columns.
    mutate(dvid = ifelse(evid == 0, 1L, NA_integer_)) |>
    tidyr::crossing(id = id_offset + seq_len(n)) |>
    select(-any_of("ID")) |>
    relocate(id) |>
    arrange(id, time, desc(evid)) |>
    mutate(group = group)
}
ev_typ_ctl <- make_events("Control", 1L)
ev_typ_rep <- make_events("Repeated dosing", 1L, id_offset = 1L)

typ <- bind_rows(
  as.data.frame(rxode2::rxSolve(mod_ctl, ev_typ_ctl, omega = NA, keep = "group")),
  as.data.frame(rxode2::rxSolve(mod_rep, ev_typ_rep, omega = NA, keep = "group"))
)

closed_form_cl <- function(ini) {
  fub <- est(ini, "fup") / est(ini, "bpr")
  sf <- est(ini, "sf_ps")
  ps_upt <- sf * (exp(est(ini, "lps_inf_act")) + exp(est(ini, "lps_dif")))
  ps_eff <- sf * exp(est(ini, "lps_dif"))
  clm <- exp(est(ini, "lcl_int_met"))
  clint_all <- ps_upt * clm / (ps_eff + clm)
  qh <- est(ini, "q_liver")
  u <- fub * clint_all / 5
  e_unit <- u / (qh + u)
  qh * (1 - (1 - e_unit)^5) * est(ini, "bpr") * 1000 # mL/h/kg, plasma
}

auc_typ <- typ |>
  group_by(group) |>
  summarise(
    auc = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    c_last = last(Cc),
    .groups = "drop"
  ) |>
  mutate(
    cl_sim = 3 / auc * 1000, # mL/h/kg
    cl_closed = c(closed_form_cl(ini_ctl), closed_form_cl(ini_rep)),
    cl_observed = c(30.02, 137.9) # Table 1
  )
knitr::kable(auc_typ, digits = 3)
stopifnot(
  # Same parameters on both sides; the gap is trapezoid error and the
  # unintegrated tail beyond 96 h only.
  all(abs(auc_typ$cl_sim / auc_typ$cl_closed - 1) < 0.02)
)
```

At the median inputs the model's plasma clearance is about 32 mL/h/kg in
the control group (observed 30.0) and about 115 mL/h/kg after repeated dosing
(observed 137.9). This reproduces the 4- to 5-fold induction of clearance that
the paper attributes to hepatic uptake.

# Monte Carlo simulation (Figure 8)

The paper drew 1000 virtual rats per group from lognormal distributions on
PSinf,act, PSdiff, CLint,met and fu,hepa (Section 2.11). Here 200 per group
are drawn from the same distributions through the models' etas.

```{r mc}
rxode2::rxSetSeed(20221122)
n_mc <- 200L
fig_grid <- seq(0, 10, by = 0.1)
make_mc_events <- function(group, n, id_offset = 0L) {
  rxode2::et(amt = 3, cmt = "central") |>
    rxode2::et(fig_grid, cmt = "central") |>
    as.data.frame() |>
    # Two `~` endpoints (Cc, Cliver): observation rows carry dvid = 1; the
    # forward solve still returns both concentration columns.
    mutate(dvid = ifelse(evid == 0, 1L, NA_integer_)) |>
    tidyr::crossing(id = id_offset + seq_len(n)) |>
    select(-any_of("ID")) |>
    relocate(id) |>
    arrange(id, time, desc(evid)) |>
    mutate(group = group)
}
mc <- bind_rows(
  as.data.frame(rxode2::rxSolve(mod_ctl, make_mc_events("Control", n_mc), keep = "group")),
  as.data.frame(rxode2::rxSolve(mod_rep, make_mc_events("Repeated dosing", n_mc, n_mc), keep = "group"))
)
stopifnot(!anyNA(mc$Cc), !anyNA(mc$Cliver))

mc_q <- mc |>
  filter(time > 0) |>
  pivot_longer(c(Cc, Cliver), names_to = "output", values_to = "conc") |>
  mutate(output = ifelse(output == "Cc", "Plasma", "Liver (hepatocellular)")) |>
  group_by(group, output, time) |>
  summarise(
    q05 = quantile(conc, 0.05), q20 = quantile(conc, 0.20),
    q35 = quantile(conc, 0.35), q50 = median(conc),
    q65 = quantile(conc, 0.65), q80 = quantile(conc, 0.80),
    q95 = quantile(conc, 0.95),
    .groups = "drop"
  )
```

```{r figure-8, fig.height = 6, fig.cap = "Replicates Figure 8 of Kim 2022: 5th-95th, 20th-80th and 35th-65th percentile bands and the median of simulated plasma (top) and hepatocellular (bottom) SCR430 after 3 mg/kg IV."}
ggplot(mc_q, aes(time)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), fill = "grey85") +
  geom_ribbon(aes(ymin = q20, ymax = q80), fill = "grey65") +
  geom_ribbon(aes(ymin = q35, ymax = q65), fill = "grey45") +
  geom_line(aes(y = q50), linetype = "dashed", colour = "darkorange", linewidth = 0.8) +
  facet_grid(output ~ group) +
  scale_y_log10(limits = c(0.01, 100)) +
  labs(x = "Time (h)", y = "SCR430 (ug/mL)")
```

Read off Figure 8, the paper's median curves run from about 25 to 4 ug/mL
(control plasma) and 22 to 0.45 ug/mL (repeated-dosing plasma) between the
first sample and 10 h. The hepatocellular medians peak near 7 ug/mL at 1 h
(control) and near 8 to 9 ug/mL within the first half hour (repeated dosing),
ending near 4 and 0.45 ug/mL at 10 h. The simulated percentiles at selected
times are below. The early repeated-dosing hepatocellular median (about
13 ug/mL at 0.5 h) runs higher than the figure's; the plasma medians and the
10 h values agree within the reading precision of the log-scale figure.

```{r fig8-numbers}
mc_q |>
  filter(time %in% c(0.5, 1, 2, 10)) |>
  select(group, output, time, q05, q50, q95) |>
  arrange(output, group, time) |>
  knitr::kable(digits = 2)
med10 <- mc_q |> filter(time == 10)
pl_ctl <- med10$q50[med10$group == "Control" & med10$output == "Plasma"]
pl_rep <- med10$q50[med10$group == "Repeated dosing" & med10$output == "Plasma"]
stopifnot(
  # Median plasma at 10 h within a factor of 1.5 of the Figure 8 median
  # (read off a log axis: about 4 and 0.45 ug/mL).
  pl_ctl > 4 / 1.5, pl_ctl < 4 * 1.5,
  pl_rep > 0.45 / 1.5, pl_rep < 0.45 * 1.5,
  # The induction effect: an order of magnitude between the groups at 10 h.
  pl_ctl / pl_rep > 5
)
```

# PKNCA validation against Table 1

The paper's non-compartmental analysis (PKSolver) used plasma sampled to 10 h.
The schedule below approximates the Figure 2a sampling times; AUC0-inf is
extrapolated from the terminal phase, as in the paper.

```{r pknca}
nca_times <- c(0, 0.083, 0.25, 0.5, 1, 2, 3, 4, 7, 10)
make_nca_events <- function(group, n, id_offset = 0L) {
  rxode2::et(amt = 3, cmt = "central") |>
    rxode2::et(nca_times, cmt = "central") |>
    as.data.frame() |>
    # Two `~` endpoints (Cc, Cliver): observation rows carry dvid = 1; the
    # forward solve still returns both concentration columns.
    mutate(dvid = ifelse(evid == 0, 1L, NA_integer_)) |>
    tidyr::crossing(id = id_offset + seq_len(n)) |>
    select(-any_of("ID")) |>
    relocate(id) |>
    arrange(id, time, desc(evid)) |>
    mutate(group = group)
}
sim_nca <- bind_rows(
  as.data.frame(rxode2::rxSolve(mod_ctl, make_nca_events("Control", n_mc), keep = "group")),
  as.data.frame(rxode2::rxSolve(mod_rep, make_nca_events("Repeated dosing", n_mc, n_mc), keep = "group"))
) |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, group) |>
  distinct(id, group, time, .keep_all = TRUE) |>
  arrange(id, time)

dose_df <- sim_nca |> distinct(id, group) |> mutate(time = 0, dose = 3)

conc_obj <- PKNCAconc(sim_nca, Cc ~ time | group + id, concu = "ug/mL", timeu = "h")
dose_obj <- PKNCAdose(dose_df, dose ~ time | group + id, route = "intravascular", doseu = "mg/kg")
intervals <- data.frame(
  start = 0, end = Inf,
  auclast = FALSE, aucinf.obs = TRUE, cl.obs = TRUE, vss.obs = TRUE, half.life = TRUE
)
nca <- pk.nca(PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_res <- as.data.frame(nca$result)
```

Table 1 reports AUC in ug/mL*min (6333 and 1420); dividing by 60 gives
105.6 and 23.7 ug*h/mL.

```{r nca-compare}
ref <- data.frame(
  group = c("Control", "Repeated dosing"),
  aucinf.obs = c(6333, 1420) / 60, # Table 1, ug/mL*min -> ug*h/mL
  cl.obs = c(30.02, 137.9) / 1000, # Table 1, mL/h/kg -> L/h/kg
  vss.obs = c(286.4, 378.4) / 1000 # Table 1, mL/kg -> L/kg
)
cmp <- ncaComparisonTable(
  simulated = nca_res,
  reference = ref,
  by = "group",
  params = c("aucinf.obs", "cl.obs", "vss.obs"),
  units = c(aucinf.obs = "ug*h/mL", cl.obs = "L/h/kg", vss.obs = "L/kg")
)
knitr::kable(cmp)

nca_med <- nca_res |>
  filter(PPTESTCD %in% c("aucinf.obs", "cl.obs")) |>
  group_by(group, PPTESTCD) |>
  summarise(med = median(PPORRES, na.rm = TRUE), .groups = "drop")
auc_ratio <- nca_med$med[nca_med$group == "Control" & nca_med$PPTESTCD == "aucinf.obs"] /
  nca_med$med[nca_med$group == "Repeated dosing" & nca_med$PPTESTCD == "aucinf.obs"]
auc_ratio
stopifnot(
  # Observed control-to-repeated AUC ratio is 4.5 (Table 1); the model's
  # median ratio must land in the same range.
  auc_ratio > 3, auc_ratio < 6,
  # Median clearance within 30% of the Table 1 group means.
  abs(nca_med$med[nca_med$group == "Control" & nca_med$PPTESTCD == "cl.obs"] / 0.03002 - 1) < 0.3,
  abs(nca_med$med[nca_med$group == "Repeated dosing" & nca_med$PPTESTCD == "cl.obs"] / 0.1379 - 1) < 0.3
)
```

The simulated medians land within about 20% of the observed group means for
AUC and clearance in both groups, and within about 12% for Vss. Vss was not a
fitting target; only the plasma profile was fitted.

# Assumptions and deviations

* **Two files, one structure.** The paper runs one mPBPK structure with two
  sets of hepatocyte-derived inputs (Table 3). The volume and exchange rate
  constants are "Equal to Control". The two sets are packaged as two files so
  that each carries its own variability magnitudes. A covariate to switch
  between them is not needed.
* **Systemic and single-adjustment equations.** Appendix A.2 prints only the
  liver equations. The systemic compartment's mass balance is read from the
  Figure 7 schematic: hepatic blood flow leaves into Inlet 1 and returns from
  Inlet 5, and `kin` / `kout` act first-order on the systemic and
  single-adjustment amounts. Because `kin` / `kout` act on amounts, the
  single-adjustment compartment needs no volume. That is consistent with
  Table 3 listing none, although Section 3.7 says that compartment's volume
  was fitted.
* **Blood vs plasma.** Eq. A1 multiplies the systemic concentration by the
  hepatic BLOOD flow, and uptake acts on `fu,b * Che`. The systemic
  compartment therefore holds blood concentration, and plasma is blood / Rb,
  using each group's Table 1 Rb. With this reading the model's steady-state
  volume is about 0.30 L/kg (control; see the PKNCA table), against the
  observed Vdss of 0.29 L/kg. Treating the systemic compartment as plasma
  would give about 0.1 L/kg.
* **Liver output.** `Cliver` is the mean hepatocellular concentration. The
  paper measured liver concentrations after correcting for blood left in the
  tissue (Eq. 1), and the hepatocellular mean reproduces the Figure 8c/8d
  median curves. Averaging the extracellular blood units in as well gives
  values about twice as high.
* **The 1.6-fold scaling factor** is applied to both PSinf,act and PSdiff on
  top of the Table 3 values, per Section 3.7. The Table 3 values sum to the
  unscaled Method 2 PSinf (see the IVIVE check above). Without the factor, the
  repeated-dosing clearance would be about 75 mL/h/kg against the observed
  137.9.
* **Text vs Table 3 for PSdiff.** Section 3.7 states PSdiff as 0.8 and
  0.6 L/h/kg. Those are the standard deviations of PSinf,act in Table 3. The
  Table 3 PSdiff means (0.3 and 0.86 L/h/kg) are used because they reproduce
  the Method 2 PSinf.
* **CLint,met (control).** Table 3 lists 9.8 L/h/kg, while Table 2's
  160.4 mL/min/kg converts to 9.6 L/h/kg. The Table 3 value is the mPBPK
  input and is used.
* **Lognormal centring.** The paper's Monte Carlo distributions are
  parameterised by their arithmetic mean (Eqs. 17 to 19). The typical values
  are the medians of those distributions, so the etas reproduce the paper's
  Monte Carlo draw exactly. A typical-value (no-eta) solve therefore uses the
  median, below the Table 3 mean by up to 28 percent (the largest gap is
  fu,hepa in the repeated-dosing group; PSinf,act and CLint,met there differ by
  well under 1 percent). The eta variances are the Eq. 18
  values and are held constant, because they come from in vitro assay
  variability and were not estimated.
* **No residual error** is reported; both proportional residual SDs are
  fixed at zero.
* **Vehicle group and intraperitoneal data** (Table 1, Figure 2b) were not
  modelled by the paper and are not part of either file.
