---
title: "Piperacillin + tazobactam in Korean patients (Kim 2016)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Piperacillin + tazobactam in Korean patients (Kim 2016)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Kim 2016 fitted piperacillin and tazobactam **separately**, applying the same
model-building process to each ("All population pharmacokinetic model processes
were equally applied for piperacillin and tazobactam", Methods section 4) and
reporting the two sets of estimates side by side in Table 2. The two fits are
therefore packaged as two independent model files, described together here.

```{r models}
ui_pip <- rxode2::rxode(readModelDb("Kim_2016_piperacillin"))
ui_taz <- rxode2::rxode(readModelDb("Kim_2016_tazobactam"))
```

* Citation: `r ui_pip$reference`
* Piperacillin model: `r ui_pip$description`
* Tazobactam model: `r ui_taz$description`
* Article: <https://doi.org/10.3947/ic.2016.48.3.209> (open access)

# Population

Thirty-three adult inpatients with acute infections were enrolled at six
university-affiliated hospitals in Korea between April 2013 and April 2015
(Methods section 1). Thirty-five patients completed the study; two were excluded
because sampling at the injection site was suspected -- their piperacillin and
tazobactam concentrations were roughly 1,000-fold higher than everyone else's
(Results section 1).

The cohort was elderly and renally heterogeneous (Table 1): age
68.79 +/- 10.97 years (range 46-88), body weight 58.17 +/- 10.08 kg (range
36.30-75.40), serum creatinine 1.18 +/- 0.80 mg/dL (range 0.34-4.70) and
Cockcroft-Gault creatinine clearance 61.27 +/- 36.67 mL/min (range
14.45-146.01). Seventeen patients (51.5%) were men. Infection sites were lungs
(13), urinary tract (12), soft tissue (6) and bloodstream (2). APACHE II score
was 13.48 +/- 8.48 (range 3-38) and Glasgow Coma Scale score 13.09 +/- 2.88
(range 7-15).

Dosing followed a renal-function rule (Methods section 2): piperacillin /
tazobactam 2 / 0.25 g for creatinine clearance <= 50 mL/min (14 patients, 42.4%)
or 4 / 0.5 g for creatinine clearance > 50 mL/min (19 patients, 57.6%), each
infused intravenously over 1 h every 8 h for at least four consecutive doses.
Blood was drawn at steady state -- pre-dose and at 0 min, 30 min and 4-6 h after
the end of the fourth infusion.

The same information is available programmatically as
`readModelDb("Kim_2016_piperacillin")()$population`.

# Source trace

Every `ini()` entry in `inst/modeldb/specificDrugs/Kim_2016_piperacillin.R` and
`inst/modeldb/specificDrugs/Kim_2016_tazobactam.R` carries an in-file comment
naming its source location. They are collected here for review.

| Equation / parameter | Piperacillin | Tazobactam | Source location |
|---|---|---|---|
| Structural model | 2-compartment, first-order elimination | 2-compartment, first-order elimination | Results sections 2 and 3; Abstract |
| `TVCL = theta1 + theta5 * CLcr/47` (L/h) | `2.90 + 4.03 * CLcr/47` | `1.76 + 4.81 * CLcr/47` | Results sections 2 and 3 ("the final model was as follows"); Table 2 structural-model header |
| `lcl_nonren` (theta1, L/h) | 2.90 (RSE 42.41%; bootstrap 2.80, 95% CI 0.01-5.26) | 1.76 (RSE 51.82%; bootstrap 1.72, 95% CI 0.01-3.67) | Table 2 |
| `lcl_renal` (theta5, L/h at CLcr = 47) | 4.03 (RSE 32.75%; bootstrap 3.99, 95% CI 1.82-7.89) | 4.81 (RSE 23.49%; bootstrap 4.79, 95% CI 2.86-7.71) | Table 2 |
| `TVV1 = theta2 * weight/60` (L) | `19.50 * weight/60` | `22.6 * weight/60` | Results sections 2 and 3; Table 2 structural-model header |
| `lvc` (theta2, L at 60 kg) | 19.50 (RSE 14.82%; bootstrap 19.31, 95% CI 15.15-29.62) | 22.6 (RSE 10.84%; bootstrap 24.10, 95% CI 20.06-30.79) | Table 2 |
| `lq` (Q, L/h) -- fixed | 2.29 (no RSE, no bootstrap) | 1.18 (no RSE, no bootstrap) | Table 2 |
| `lvp` (V2, L) -- fixed | 3.76 (no RSE, no bootstrap) | 4.3 (no RSE, no bootstrap) | Table 2 |
| `crcl_ref` (mL/min) | 47 | 47 | Printed in the Table 2 / Results CL equation |
| `wt_ref` (kg) | 60 | 60 | Results sections 2 and 3 ("median body weight of 60 kg") |
| `etalcl` (omega CL) | 0.279 -> variance 0.077841 | 0.211 -> variance 0.044521 | Table 2, "Random variability (CV,%)" |
| `etalvc` (omega V1) | 0.179 -> variance 0.032041 | 0.151 -> variance 0.022801 | Table 2, "Random variability (CV,%)" |
| `propSd` (sigma, proportional) | 0.399 (RSE 10.78%; bootstrap 0.390, 95% CI 0.297-0.472) | 0.393 (RSE 12.26%; bootstrap 0.387, 95% CI 0.297-0.479) | Table 2, "Residual variability (%)" |
| Residual-error form | proportional only | proportional only | Results sections 2 and 3 |
| IIV distribution | log-normal, independent etas | log-normal, independent etas | Methods section 4 |
| Covariate CLcr on CL | dOFV -10.387 | dOFV -17.736 | Results sections 2 and 3 |
| Covariate weight on V1 | a priori, linear | a priori, linear | Methods section 4 (following the paper's reference [16]) |
| Input | 1-h zero-order IV infusion into `central` | 1-h zero-order IV infusion into `central` | Methods section 2 |

# Virtual cohort

The observed concentrations are not publicly available, so a virtual cohort is
drawn whose covariate distributions approximate the Table 1 demographics and
whose dose assignment follows the study's renal-function rule exactly.

```{r cohort}
# set.seed() seeds R's RNG (used for the covariate draw below). It does NOT
# seed rxode2's simulation RNG, whose streams are partitioned per solver
# thread -- so the eta draw differs between a 2-core CI runner and a
# multi-thread workstation. Every assertion downstream is written to hold for
# any cohort this model can produce (see the deterministic gates further down,
# which use zeroRe() and therefore do not depend on the eta draw at all).
set.seed(20160716)

n_per_arm <- 150L

# Truncated normal draw: Table 1 gives mean +/- SD and an observed range for
# each continuous covariate, so values are drawn from the normal and resampled
# until they fall inside the reported range.
rtrunc_norm <- function(n, mean, sd, lo, hi) {
  out <- numeric(0)
  while (length(out) < n) {
    x <- stats::rnorm(n * 4L, mean, sd)
    out <- c(out, x[x >= lo & x <= hi])
  }
  out[seq_len(n)]
}

# Table 1: body weight 58.17 +/- 10.08 kg, range 36.30-75.40.
# Table 1: CLcr 61.27 +/- 36.67 mL/min, range 14.45-146.01. Methods section 2
# splits the cohort at 50 mL/min, so each arm is drawn from its own side of the
# split; this reproduces the protocol and fixes the arm sizes.
make_cohort <- function(n, crcl_lo, crcl_hi, treatment, id_offset = 0L) {
  tibble::tibble(
    id = id_offset + seq_len(n),
    WT = rtrunc_norm(n, 58.17, 10.08, 36.30, 75.40),
    CRCL = rtrunc_norm(n, 61.27, 36.67, crcl_lo, crcl_hi),
    treatment = treatment
  )
}

cohort <- dplyr::bind_rows(
  make_cohort(n_per_arm, 14.45, 50.00, "2/0.25 g q8h (CLcr <= 50)", id_offset = 0L),
  make_cohort(n_per_arm, 50.00, 146.01, "4/0.5 g q8h (CLcr > 50)", id_offset = n_per_arm)
)

stopifnot(
  nrow(cohort) == 2L * n_per_arm,
  !anyDuplicated(cohort$id),
  all(cohort$WT >= 36.30 & cohort$WT <= 75.40),
  all(cohort$CRCL >= 14.45 & cohort$CRCL <= 146.01)
)
```

The study sampled after the fourth infusion. Eight q8h doses are simulated here
instead, so that the last dosing interval is at steady state to within numerical
noise even for the slowest-clearing subject; the NCA and the closed-form gates
below all run over that last interval.

```{r events}
tau <- 8 # dosing interval (h), Methods section 2
n_doses <- 8L
dose_times <- (seq_len(n_doses) - 1L) * tau
ss_start <- max(dose_times) # start of the final (steady-state) dosing interval
ss_end <- ss_start + tau

# Coarse grid over the approach to steady state, dense grid over the final
# interval so the trapezoidal AUC resolves the 1-h infusion peak.
obs_times <- c(
  seq(0, ss_start - 2, by = 2),
  seq(ss_start, ss_end, by = 0.05)
)

# Piperacillin 2 g / 4 g, tazobactam 0.25 g / 0.5 g (Methods section 2),
# infused over 1 h. `cmt` is the ODE state name on BOTH dose and observation
# rows -- never the algebraic observable `Cc`.
make_events <- function(amt_low, amt_high) {
  doses <- cohort |>
    tidyr::crossing(time = dose_times) |>
    dplyr::mutate(
      evid = 1L,
      cmt = "central",
      amt = ifelse(treatment == "2/0.25 g q8h (CLcr <= 50)", amt_low, amt_high),
      dur = 1
    )
  obs <- cohort |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(evid = 0L, cmt = "central", amt = NA_real_, dur = NA_real_)
  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

ev_pip <- make_events(2000, 4000)
ev_taz <- make_events(250, 500)

stopifnot(
  !anyDuplicated(unique(ev_pip[, c("id", "time", "evid")])),
  !anyDuplicated(unique(ev_taz[, c("id", "time", "evid")]))
)
```

# Simulation

```{r simulate}
mod_pip <- readModelDb("Kim_2016_piperacillin")
mod_taz <- readModelDb("Kim_2016_tazobactam")

keep_cols <- c("treatment", "WT", "CRCL")

sim_pip <- rxode2::rxSolve(mod_pip, events = ev_pip, keep = keep_cols) |>
  as.data.frame()
sim_taz <- rxode2::rxSolve(mod_taz, events = ev_taz, keep = keep_cols) |>
  as.data.frame()

stopifnot(
  nrow(sim_pip) > 0, nrow(sim_taz) > 0,
  all(sim_pip$Cc >= 0, na.rm = TRUE),
  all(sim_taz$Cc >= 0, na.rm = TRUE)
)
```

A typical-value (between-subject variability removed) companion is used for the
deterministic gates; with `zeroRe()` each subject's clearance is exactly the
typical value implied by their own covariates, so it can be compared against the
equation as the paper prints it.

```{r simulate-typical}
sim_pip_tv <- rxode2::rxSolve(rxode2::zeroRe(mod_pip), events = ev_pip, keep = keep_cols) |>
  as.data.frame()
sim_taz_tv <- rxode2::rxSolve(rxode2::zeroRe(mod_taz), events = ev_taz, keep = keep_cols) |>
  as.data.frame()
```

# Replicate published figures

Kim 2016 Figure 2 is a visual predictive check for the final piperacillin model:
2,000 simulated patients, with the median and the 2.5th / 97.5th predicted
percentiles overlaid on the observed concentrations. The observed data are not
available here, so the panel below shows the simulated prediction interval only,
for the steady-state dosing interval that the study sampled. (Methods section 4
describes the VPC as using 5th / 50th / 95th percentile curves while the
Figure 2 caption describes 2.5th / 97.5th; the caption is followed here.)

```{r figure-2}
# Replicates Figure 2 of Kim 2016 (piperacillin VPC), plus the same view for
# tazobactam, over the final steady-state dosing interval.
vpc_band <- function(sim, drug) {
  sim |>
    dplyr::filter(time >= ss_start, time <= ss_end) |>
    dplyr::mutate(tad = time - ss_start, drug = drug) |>
    dplyr::group_by(drug, treatment, tad) |>
    dplyr::summarise(
      Q025 = stats::quantile(Cc, 0.025, na.rm = TRUE),
      Q50 = stats::quantile(Cc, 0.50, na.rm = TRUE),
      Q975 = stats::quantile(Cc, 0.975, na.rm = TRUE),
      .groups = "drop"
    )
}

dplyr::bind_rows(
  vpc_band(sim_pip, "Piperacillin"),
  vpc_band(sim_taz, "Tazobactam")
) |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q025, ymax = Q975), alpha = 0.25) +
  geom_line() +
  facet_grid(drug ~ treatment, scales = "free_y") +
  labs(
    x = "Time after the start of the steady-state infusion (h)",
    y = "Plasma concentration (mg/L)",
    title = "Steady-state prediction intervals (median, 2.5th-97.5th percentile)",
    caption = "Replicates the simulated bands of Figure 2 of Kim 2016; observed points unavailable."
  ) +
  theme_bw()
```

The Discussion contrasts the Korean piperacillin clearance equation with one
developed in Caucasians, `CL = (1.29 * CLcr + 100) * 60/1000` L/h, and concludes
that "clearance of piperacillin in Korean patients is expected to be lower than
that in Caucasians in both normal and lower renal function within the normal
body weight range". That comparison is reproduced below.

```{r figure-discussion}
crcl_grid <- seq(14.45, 146.01, length.out = 200)
cl_compare <- dplyr::bind_rows(
  tibble::tibble(
    CRCL = crcl_grid,
    CL = 2.90 + 4.03 * crcl_grid / 47,
    Equation = "Korean (Kim 2016)"
  ),
  tibble::tibble(
    CRCL = crcl_grid,
    CL = (1.29 * crcl_grid + 100) * 60 / 1000,
    Equation = "Caucasian (Kim 2016 Discussion)"
  )
)

ggplot(cl_compare, aes(CRCL, CL, colour = Equation)) +
  geom_line(linewidth = 0.9) +
  labs(
    x = "Creatinine clearance (mL/min)",
    y = "Typical piperacillin CL (L/h)",
    title = "Korean vs Caucasian piperacillin clearance equations",
    caption = "Reproduces the comparison made in the Discussion of Kim 2016."
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

# PKNCA validation

NCA is computed over the final steady-state dosing interval, grouped by
treatment arm.

```{r pknca}
# Only `!is.na(Cc)` -- adding `time > 0` or `Cc > 0` would drop the row that
# anchors the AUC interval.
nca_conc <- function(sim) {
  sim |>
    dplyr::filter(!is.na(Cc), time >= ss_start, time <= ss_end) |>
    dplyr::select(id, time, Cc, treatment)
}

nca_dose <- function(ev) {
  ev |>
    dplyr::filter(evid == 1, time == ss_start) |>
    dplyr::select(id, time, amt, treatment)
}

ss_intervals <- data.frame(
  start = ss_start,
  end = ss_end,
  cmax = TRUE,
  tmax = TRUE,
  cmin = TRUE,
  auclast = TRUE,
  cav = TRUE
)

run_nca <- function(sim, ev, concu, doseu) {
  conc_obj <- PKNCA::PKNCAconc(
    nca_conc(sim), Cc ~ time | treatment + id,
    concu = concu, timeu = "h"
  )
  dose_obj <- PKNCA::PKNCAdose(
    nca_dose(ev), amt ~ time | treatment + id,
    doseu = doseu
  )
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = ss_intervals))
}

nca_pip <- run_nca(sim_pip, ev_pip, "mg/L", "mg")
nca_taz <- run_nca(sim_taz, ev_taz, "mg/L", "mg")

stopifnot(
  nrow(as.data.frame(nca_pip$result)) > 0,
  nrow(as.data.frame(nca_taz$result)) > 0
)
```

```{r pknca-table}
nca_summary <- function(res, drug) {
  as.data.frame(res$result) |>
    dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "cmin", "auclast", "cav")) |>
    dplyr::group_by(treatment, PPTESTCD) |>
    dplyr::summarise(
      median = stats::median(PPORRES),
      q05 = stats::quantile(PPORRES, 0.05),
      q95 = stats::quantile(PPORRES, 0.95),
      .groups = "drop"
    ) |>
    dplyr::mutate(
      Drug = drug,
      Parameter = nlmixr2lib::ncaParamLabel(PPTESTCD),
      Value = sprintf("%.3g (%.3g, %.3g)", median, q05, q95)
    ) |>
    dplyr::select(Drug, treatment, Parameter, Value)
}

dplyr::bind_rows(
  nca_summary(nca_pip, "Piperacillin"),
  nca_summary(nca_taz, "Tazobactam")
) |>
  dplyr::rename("Dose arm" = treatment) |>
  knitr::kable(
    caption = paste(
      "Simulated steady-state NCA over the final dosing interval:",
      "median (5th, 95th percentile) across 150 subjects per arm.",
      "Concentrations in mg/L, AUC in mg*h/L, times in h."
    )
  )
```

## Comparison against published NCA

Kim 2016 reports **no** non-compartmental parameters -- neither a Cmax / Tmax /
AUC / half-life table nor per-arm exposure summaries. Figure 1 is a
goodness-of-fit panel and Figure 2 a visual predictive check; Table 1 is
demographics and Table 2 is model parameters. There is therefore nothing to put
on the reference side of `nlmixr2lib::ncaComparisonTable()`, and no such table is
rendered. The closed-form gates below take its place: they check that the
packaged model reproduces the equations the paper *does* print, and that the
simulated exposures are internally consistent with them.

# Closed-form validation gates

## Gate 1 -- steady-state AUC reproduces the paper's printed clearance equation

At steady state, the AUC over one dosing interval satisfies
`AUC_tau = Dose / CL` exactly, for any linear multi-compartment model. The
clearance implied by the simulated profile is therefore `Dose / AUC_tau`, and it
must equal the clearance obtained by evaluating the paper's printed equation at
the subject's own creatinine clearance. The equations are transcribed here
directly from Results sections 2 and 3 -- independently of how the model file
encodes them -- so a mis-transcribed intercept, slope or reference constant in
the model file makes this gate go red.

```{r gate-1}
nca_typical <- function(sim, ev, concu) {
  res <- run_nca(sim, ev, concu, "mg")
  as.data.frame(res$result) |>
    dplyr::filter(PPTESTCD == "auclast") |>
    dplyr::select(id, treatment, auc_tau = PPORRES) |>
    dplyr::left_join(
      ev |>
        dplyr::filter(evid == 1, time == ss_start) |>
        dplyr::select(id, amt, CRCL),
      by = "id"
    ) |>
    dplyr::mutate(cl_from_auc = amt / auc_tau)
}

# Kim 2016 Results section 2: CL (L/h) = 2.9 + 4.03 * CLcr/47
gate1_pip <- nca_typical(sim_pip_tv, ev_pip, "mg/L") |>
  dplyr::mutate(
    cl_from_paper = 2.9 + 4.03 * CRCL / 47,
    pct_diff = 100 * (cl_from_auc - cl_from_paper) / cl_from_paper
  )

# Kim 2016 Results section 3: CL (L/h) = 1.76 + 4.81 * CLcr/47
gate1_taz <- nca_typical(sim_taz_tv, ev_taz, "mg/L") |>
  dplyr::mutate(
    cl_from_paper = 1.76 + 4.81 * CRCL / 47,
    pct_diff = 100 * (cl_from_auc - cl_from_paper) / cl_from_paper
  )

# Deterministic comparison: both sides use the same typical-value parameters,
# so the only difference is trapezoidal error on a 0.05 h grid plus the
# residual approach to steady state after 7 dosing intervals. A tight all()
# bound is correct here (there is no per-subject random mechanism separating
# the two sides), and it goes red on any transcription error in theta1,
# theta5 or the 47 mL/min reference -- each of which moves CL by tens of
# percent.
stopifnot(
  max(abs(gate1_pip$pct_diff)) < 1,
  max(abs(gate1_taz$pct_diff)) < 1
)

tibble::tibble(
  Drug = c("Piperacillin", "Tazobactam"),
  `Max |% difference|` = c(
    max(abs(gate1_pip$pct_diff)),
    max(abs(gate1_taz$pct_diff))
  ),
  `Median |% difference|` = c(
    stats::median(abs(gate1_pip$pct_diff)),
    stats::median(abs(gate1_taz$pct_diff))
  )
) |>
  knitr::kable(
    digits = 4,
    caption = "Gate 1: Dose / AUC_tau at steady state vs the paper's printed CL equation."
  )
```

## Gate 2 -- central volume reproduces the printed weight relationship

The end-of-infusion concentration of the *first* dose is not a clean read on V1
for a two-compartment model, but the model's own `vc` output column is, and it
must equal `theta2 * weight/60` evaluated independently from Results sections 2
and 3. This gate protects the `wt_ref = 60` reference and the linear (exponent
1, not allometric 0.75) form.

```{r gate-2}
vc_check <- function(sim_tv, theta2) {
  sim_tv |>
    dplyr::distinct(id, WT, vc) |>
    dplyr::mutate(
      vc_from_paper = theta2 * WT / 60,
      pct_diff = 100 * (vc - vc_from_paper) / vc_from_paper
    )
}

gate2_pip <- vc_check(sim_pip_tv, 19.50)
gate2_taz <- vc_check(sim_taz_tv, 22.6)

stopifnot(
  max(abs(gate2_pip$pct_diff)) < 1e-8,
  max(abs(gate2_taz$pct_diff)) < 1e-8
)

tibble::tibble(
  Drug = c("Piperacillin", "Tazobactam"),
  `V1 at 60 kg (L)` = c(
    unique(round(gate2_pip$vc_from_paper[which.min(abs(gate2_pip$WT - 60))], 2)),
    unique(round(gate2_taz$vc_from_paper[which.min(abs(gate2_taz$WT - 60))], 2))
  ),
  `Max |% difference|` = c(
    max(abs(gate2_pip$pct_diff)),
    max(abs(gate2_taz$pct_diff))
  )
) |>
  knitr::kable(caption = "Gate 2: model V1 vs the printed theta2 * weight/60 relationship.")
```

## Gate 3 -- terminal half-life reproduces the two-compartment closed form

Q and V2 are not exercised by Gate 1 (steady-state AUC depends only on CL) or
Gate 2. They are gated here through the terminal disposition rate constant of
the two-compartment system,

`beta = 0.5 * ((kel + k12 + k21) - sqrt((kel + k12 + k21)^2 - 4 * kel * k21))`,

compared against the PKNCA half-life of a single-dose washout at the reference
subject (60 kg, CLcr 47 mL/min). A wrong Q or V2 changes `beta` and the gate
goes red.

`lambda_z` is restricted to 28 h post-dose and later rather than left to PKNCA's
automatic point selection. For tazobactam the distribution rate constant
(`alpha = 0.43 /h`) is only about 2.4-fold faster than `beta = 0.18 /h`, so an
earlier window still carries distribution-phase curvature and `lambda_z` reads
steep: PKNCA's automatic selection began at 13.2 h and returned a half-life of
3.65 h against a closed-form 3.76 h -- a 3.0% bias in the *instrument*, not in
the model (see pattern 11 of the skill's known-failure catalogue). `r.squared`
does not warn about it; it was 0.9999 on the biased fit.

The window is set with `PKNCAconc(include_half.life = )`, which takes the name
of a **logical column of the concentration data**. Note that
`lambda.z.time.first` in the `intervals` data frame is a reporting flag, not a
window setting -- putting a time there is rejected outright.

```{r gate-3}
ref_subject <- tibble::tibble(
  id = 1L, WT = 60, CRCL = 47, treatment = "reference"
)

washout_end <- 40 # h of observation after the single dose
lambda_z_start <- 28 # h; start of the terminal-slope window

washout_events <- function(amt) {
  dplyr::bind_rows(
    ref_subject |> dplyr::mutate(time = 0, evid = 1L, cmt = "central", amt = amt, dur = 1),
    ref_subject |>
      tidyr::crossing(time = seq(0, washout_end, by = 0.05)) |>
      dplyr::mutate(evid = 0L, cmt = "central", amt = NA_real_, dur = NA_real_)
  ) |>
    dplyr::arrange(time, dplyr::desc(evid))
}

half_life_check <- function(mod, amt, theta1, theta5, theta2, q, vp) {
  ev <- washout_events(amt)
  sim <- rxode2::rxSolve(rxode2::zeroRe(mod), events = ev, keep = c("treatment")) |>
    as.data.frame()
  if (is.null(sim$id)) sim$id <- 1L
  stopifnot(all(sim$Cc >= 0, na.rm = TRUE))

  conc <- sim |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::select(id, time, Cc, treatment) |>
    dplyr::mutate(include_hl = time >= lambda_z_start)
  dose <- ev |>
    dplyr::filter(evid == 1) |>
    dplyr::select(id, time, amt, treatment)

  # The tail is still far above solver noise at 40 h (0.0037 mg/L for
  # piperacillin against a Cmax of 165 mg/L, 0.0056 against 18.8 for
  # tazobactam), so the log-linear fit is not reading rounding error.
  stopifnot(
    sum(conc$include_hl) > 50,
    all(conc$Cc[conc$include_hl] > 1e-4)
  )

  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(
      conc, Cc ~ time | treatment + id,
      concu = "mg/L", timeu = "h",
      include_half.life = "include_hl"
    ),
    PKNCA::PKNCAdose(dose, amt ~ time | treatment + id, doseu = "mg"),
    intervals = data.frame(start = 0, end = Inf, half.life = TRUE)
  ))
  hl_nca <- as.data.frame(res$result) |>
    dplyr::filter(PPTESTCD == "half.life") |>
    dplyr::pull(PPORRES)

  # Closed form from values transcribed directly from Kim 2016 Table 2.
  cl <- theta1 + theta5 * 47 / 47
  v1 <- theta2 * 60 / 60
  kel <- cl / v1
  k12 <- q / v1
  k21 <- q / vp
  s <- kel + k12 + k21
  beta <- 0.5 * (s - sqrt(s^2 - 4 * kel * k21))
  list(hl_nca = hl_nca, hl_closed = log(2) / beta)
}

gate3_pip <- half_life_check(mod_pip, 4000, 2.90, 4.03, 19.50, 2.29, 3.76)
gate3_taz <- half_life_check(mod_taz, 500, 1.76, 4.81, 22.6, 1.18, 4.3)

gate3 <- tibble::tibble(
  Drug = c("Piperacillin", "Tazobactam"),
  `Closed-form t1/2 (h)` = c(gate3_pip$hl_closed, gate3_taz$hl_closed),
  `PKNCA t1/2 (h)` = c(gate3_pip$hl_nca, gate3_taz$hl_nca)
) |>
  dplyr::mutate(
    `% difference` = 100 * (`PKNCA t1/2 (h)` - `Closed-form t1/2 (h)`) / `Closed-form t1/2 (h)`
  )

# Deterministic: a typical-value solve against its own closed form, with no
# per-subject random mechanism separating the two sides, so a tight bound is
# correct here rather than a cohort-robust one. Realised -0.000% (piperacillin)
# and -0.070% (tazobactam) on the 28-40 h window. 0.5% leaves headroom for
# platform-to-platform solver differences while still going red on a wrong Q or
# V2: halving V2 moves the piperacillin closed-form t1/2 by 16% and doubling Q
# moves it by 7%.
stopifnot(max(abs(gate3$`% difference`)) < 0.5)

knitr::kable(
  gate3,
  digits = 3,
  caption = "Gate 3: PKNCA terminal half-life vs the two-compartment closed form at the reference subject."
)
```

## Gate 4 -- the Discussion's Korean-vs-Caucasian clearance claim

```{r gate-4}
gate4 <- tibble::tibble(
  CRCL = c(20, 47, 100, 146),
  `Korean CL (L/h)` = 2.90 + 4.03 * CRCL / 47,
  `Caucasian CL (L/h)` = (1.29 * CRCL + 100) * 60 / 1000
) |>
  dplyr::mutate(`Korean lower?` = `Korean CL (L/h)` < `Caucasian CL (L/h)`)

# The paper's Discussion claim is categorical ("lower ... in both normal and
# lower renal function"), and both equations here are deterministic, so an
# exact comparison is appropriate -- no simulated quantity is involved.
stopifnot(all(gate4$`Korean lower?`))

knitr::kable(
  gate4,
  digits = 2,
  caption = paste(
    "Gate 4: Kim 2016 Discussion -- Korean piperacillin clearance is lower than",
    "the Caucasian equation across the observed CLcr range."
  )
)
```

# Assumptions and deviations

## Interpretation of Table 2's random-effects block

Table 2 groups the two etas under a header reading **"Random variability
(CV,%)"** and the residual under **"Residual variability (%)"**, but prints
fractions (0.279, 0.179, 0.399) rather than percentages, and labels the rows with
the symbols `omega` and `sigma`. Two readings are possible: the printed numbers
are standard deviations (the NONMEM meaning of `omega` / `sigma`), or they are
variances.

The **standard-deviation reading was adopted**, on three grounds:

1. *The residual row's RSE sits below the variance-scale floor.* The asymptotic
   relative standard error of a variance estimated from `N` effectively
   independent residuals is `sqrt(2/N)`. With 33 patients sampled four times
   each (Methods section 2), `N` is at most 132 and `sqrt(2/132) = 12.31%` is a
   floor that a variance-scale RSE column could not beat. Table 2 reports
   10.78% for piperacillin and 12.26% for tazobactam -- both below it. The
   corresponding floor for an SD is half that, `8.7%`, which both clear
   comfortably. The same test is inconclusive for the two eta rows (33
   subjects gives a floor of 24.62% and the reported RSEs are 24.73%, 28.67%,
   84.36% and 68.87%), so the residual row is what settles the block's scale.
2. *Plausibility.* Under the variance reading the proportional residual error
   would be `sqrt(0.399) = 63.2%` for piperacillin and 62.7% for tazobactam,
   which is not credible for the fully validated LC-MS/MS assay of Methods
   section 3 (calibration `R^2 > 0.995`). The SD reading gives 39.9% and 39.3%,
   high but consistent with sparse clinical sampling and recorded-dose-time
   error.
3. *Internal consistency of the header.* For a log-normal eta this small, the
   coefficient of variation is approximately equal to `omega`, so 0.279 reads
   as 27.9% CV -- which is what the "(CV,%)" header claims. Under the variance
   reading the printed number is neither a CV nor a percentage and the header
   would simply be wrong.

The model files therefore encode `etalcl ~ omega^2` and `etalvc ~ omega^2`
(piperacillin 0.077841 and 0.032041; tazobactam 0.044521 and 0.022801) and
`propSd` as the printed value. A closely related sub-reading -- that 0.279 is a
CV rather than an `omega`, giving `log(1 + 0.279^2) = 0.074988` -- differs from
the adopted variance by 3.7%, i.e. 1.9% on the SD scale, and is not separately
encoded.

## Q and V2 were treated as fixed

`Q` and `V2` are the only two structural rows in Table 2 with a dash in **both**
the `%RSE` and the bootstrap columns, for both drugs; every other row has both.
NONMEM relative standard errors come from the covariance matrix, and the
presence of RSEs on every other row proves the covariance step succeeded -- so
`Q` and `V2` would have carried RSEs too had they been estimated. They are
therefore encoded with `fixed()`. The paper prints no `FIX` flag and does not
say where the two values came from. Because the values themselves are used
either way, this choice affects only the provenance metadata, not any
simulated quantity.

## The 47 mL/min reference in the clearance equation

The clearance equation normalises creatinine clearance by 47 mL/min, in both the
Results text and the Table 2 structural-model header. The paper never states
what 47 is. Table 1 reports the cohort **mean** creatinine clearance (61.27
mL/min) and not the median, so 47 cannot be confirmed as the cohort median from
the text. It is reproduced exactly as printed. By contrast the V1 reference of
60 kg *is* stated ("adjusted for median body weight of 60 kg").

## Other assumptions

- **Covariate distributions.** Table 1 reports mean +/- SD and an observed range
  for each continuous covariate but no joint distribution. Weight and creatinine
  clearance are drawn independently from normals truncated to the reported
  ranges. The two dose arms are drawn from either side of the protocol's
  50 mL/min split (Methods section 2), which fixes the arm sizes at 150 each
  rather than reproducing the study's 14 / 19 split.
- **Number of doses.** The study sampled after the fourth infusion; eight
  infusions are simulated here so the final interval is at steady state to
  within numerical noise even for the slowest-clearing subject (terminal
  half-life at CLcr 14.45 mL/min is about 4 h, so seven intervals is roughly 14
  half-lives). This affects the gates' precision, not the model.
- **Observed data.** The individual concentrations behind Figures 1 and 2 are
  not published, so the VPC panel shows the simulated prediction interval with
  no observed overlay, and the goodness-of-fit panels of Figure 1 cannot be
  reproduced at all.
- **No published NCA to compare against.** Kim 2016 reports no non-compartmental
  parameters, so `nlmixr2lib::ncaComparisonTable()` is not used; the closed-form
  gates above validate against the equations the paper does print.
- **Glasgow Coma Scale.** GCS was screened as a candidate covariate (Methods
  section 4) and not retained. It has no canonical covariate-column name in
  `inst/references/covariate-columns.md`, and this extraction does not introduce
  one because no model uses it; it is recorded in the models'
  `population$notes` rather than in `covariatesDataExcluded`. The other screened
  and non-retained covariates (age, sex, blood urea nitrogen, APACHE II score)
  are recorded in `covariatesDataExcluded`.
- **Blood urea nitrogen.** Listed among the screened covariates in Methods
  section 4 but absent from the Table 1 demographics, so no distribution for it
  is reproduced here.
- **Non-paper-derived parameter values.** None. Every value in both model files
  comes from the Kim 2016 text or Table 2.
