---
title: "Tegafur / 5-FU with Sipjeondaebo-tang in rats (Kim 2017)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tegafur / 5-FU with Sipjeondaebo-tang in rats (Kim 2017)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Kim_2017_tegafur_rat"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Kim_2017_tegafur_rat"))$description`
* Article: <https://doi.org/10.3390/molecules22091488> (open access, PMC6151713)

No supplementary material accompanies this article; every value below comes
from the main text, Table 1 or Table 2.

# Population

Kim 2017 studied male Sprague-Dawley rats (8 weeks, 280-300 g) at the Catholic
University of Daegu (IACUC-2013-025). Rats were pretreated by oral gavage with
either Sipjeondaebo-tang (SDT) 1200 mg/kg or the 1% CMC-Na vehicle, as a single
dose or once daily for seven consecutive days (n = 5 per arm per schedule).
One minute after the final pretreatment dose, S-1 was given orally as a mixture
of tegafur 5 mg/kg, gimeracil 1.45 mg/kg and oteracil potassium 4.9 mg/kg in
7.5% DMSO. Plasma was sampled predose and at 0.25, 0.5, 1, 1.5, 2, 3, 4, 8, 12
and 24 h, and tegafur, 5-FU and gimeracil were assayed by LC/MS/MS (LLOQ 50, 10
and 50 ng/mL).

The population model (Section 4.5) was fitted to the tegafur and 5-FU profiles
after oral S-1, together with a plasma 5-FU profile after intravenous 5-FU
10 mg/kg from an earlier study by the same group. **Gimeracil was assayed and
analysed noncompartmentally but is not part of the population model**, so it is
not represented in this model file; the gimeracil result is nevertheless the
paper's mechanistic explanation for the 5-FU effect.

The same information is available programmatically via
`readModelDb("Kim_2017_tegafur_rat")()$population`.

# Model structure

The model is a parent-metabolite cascade with a **branched absorption site**
(Kim 2017 Figure 2 and equations 1-6). The dosing compartment drains by two
competing first-order routes:

* `ka` carries intact tegafur into a two-compartment tegafur disposition model;
* `k_precursor_5fu_form` (the paper's `Ka,Met`) represents first-pass
  metabolism in gut and liver, delivering drug directly into an amount-only
  pool of the 5-FU precursor 5'-hydroxytegafur.

The precursor pool is additionally fed *systemically* by the fraction `fm`
(the paper's `FMet`) of tegafur clearance, and converts to 5-FU at
`k_5fu_form` (`KConv`). 5-FU then follows its own two-compartment disposition.

Because `ka` (0.296 1/h) is far below the tegafur elimination rate constant
`cl/vc` (0.0813 / 0.0464 = 1.75 1/h), tegafur is **flip-flop**: the terminal
slope reflects absorption, not elimination. The paper uses this to explain why
the 5-FU half-life (2.2-3.1 h) tracks tegafur's rather than 5-FU's own
intrinsic half-life of under 20 minutes.

SDT pretreatment enters as the binary covariate `CONMED_SDT`, which selects
between separately estimated control and SDT values of `ka`,
`k_precursor_5fu_form` and `cl_5fu`. All other parameters are shared.

# Source trace

Every `ini()` entry carries an in-file comment naming its source row; they are
collected here for review.

| Equation / parameter | Value | Source location |
|---|---|---|
| `d/dt(depot)` | n/a | Kim 2017 equation 1 |
| `d/dt(central)` | n/a | Kim 2017 equation 2 |
| `d/dt(peripheral1)` | n/a | Kim 2017 equation 3 |
| `d/dt(precursor_5fu)` | n/a | Kim 2017 equation 4 |
| `d/dt(central_5fu)` | n/a | Kim 2017 equation 5 |
| `d/dt(peripheral1_5fu)` | n/a | Kim 2017 equation 6 |
| `lka_ctl` (Ka,Con) | 0.296 1/h | Table 2 row 1 |
| `lka_sdt` (Ka,Pretre) | 0.197 1/h | Table 2 row 2 |
| `lk_precursor_5fu_form_ctl` (Ka,Met,Con) | 0.122 1/h | Table 2 row 3 |
| `lk_precursor_5fu_form_sdt` (Ka,Met,Pretre) | 0.0595 1/h | Table 2 row 4 |
| `lk_5fu_form` (KConv) | 2.88 1/h | Table 2 row 5 |
| `lcl` (CLTeg/F) | 0.0813 L/h/kg | Table 2 row 6 |
| `fm` (FMet) | 0.342 | Table 2 row 7 |
| `lcl_5fu_ctl` (CL5FU,Con/F) | 3.52 L/h/kg | Table 2 row 8 |
| `lcl_5fu_sdt` (CL5FU,Pretre/F) | 5.93 L/h/kg | Table 2 row 9 |
| `lq` (CLdTeg/F) | 0.184 L/h/kg | Table 2 row 10 |
| `lq_5fu` (CLd5FU/F) | 1.87 L/h/kg | Table 2 row 11 |
| `lvc` (V1,Teg/F) | 0.0464 L/kg | Table 2 row 12 |
| `lvc_5fu` (V1,5FU/F) | 0.623 L/kg | Table 2 row 13 |
| `lvp` (V2,Teg/F) | 0.137 L/kg | Table 2 row 14 |
| `lvp_5fu` (V2,5FU/F) | 0.294 L/kg | Table 2 row 15 |
| All `eta*` variances | see Table 2 | Table 2 parenthesised BSV column |
| Residual error | not published | Section 4.5 declares the structure only |
| Dose 5 mg/kg tegafur | 5000 ug/kg | Section 4.2 |
| Sampling schedule | 0-24 h | Section 4.2 |

# Units

Amounts are carried in ug/kg and volumes in L/kg (the paper's clearances and
volumes are body-weight normalised), so concentrations come out in ug/L, which
is numerically ng/mL -- the unit Table 1 uses. The tegafur dose of 5 mg/kg is
therefore entered as `amt = 5000`.

```{r units-check}
mod <- readModelDb("Kim_2017_tegafur_rat")
ui <- rxode2::rxode2(mod)
stopifnot(identical(ui$meta$units$concentration, "ng/mL"))
```

# Virtual cohort

Original observed data are not public. Two arms of 200 virtual rats each are
simulated at the paper's actual sampling times, so the NCA below uses the same
trapezoidal support the published analysis did.

```{r cohort}
# rxode2's RNG is partitioned per solver thread, so this cohort is not
# byte-identical across machines with different thread counts. Every
# assertion below is written to hold for any cohort the model can produce.
set.seed(20170907)
rxode2::rxSetSeed(20170907)

sample_times <- c(0, 0.25, 0.5, 1, 1.5, 2, 3, 4, 8, 12, 24)
n_per_arm <- 200L
tegafur_dose_ug_kg <- 5000

make_arm <- function(n, sdt, label, id_offset = 0L) {
  ids <- id_offset + seq_len(n)
  dosing <- tibble(
    id = ids, time = 0, amt = tegafur_dose_ug_kg,
    evid = 1L, cmt = "depot"
  )
  # Two endpoints are declared in the model, so observation rows must name
  # the OBSERVABLE (not the ODE state). Cc and Cc_5fu are both returned as
  # columns at every output row, so one set of rows serves both analytes.
  obs <- tidyr::expand_grid(id = ids, time = sample_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "Cc")
  bind_rows(dosing, obs) |>
    mutate(CONMED_SDT = sdt, arm = label) |>
    arrange(id, time, desc(evid))
}

events <- bind_rows(
  make_arm(n_per_arm, 0, "Control", id_offset = 0L),
  make_arm(n_per_arm, 1, "SDT", id_offset = n_per_arm)
)

# Disjoint IDs across arms are mandatory: rxSolve keys subjects on id and
# would silently merge duplicates into one double-dosed subject.
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
stopifnot(length(intersect(
  events$id[events$arm == "Control"], events$id[events$arm == "SDT"]
)) == 0)
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(mod, events = events, keep = c("arm", "CONMED_SDT")) |>
  as.data.frame()

# Residual error is fixed(0) in this model (not published), so Cc / Cc_5fu are
# the individual predictions; they are used directly rather than `sim`.
stopifnot(all(c("Cc", "Cc_5fu", "arm") %in% names(sim)))
```

A deterministic typical-value replication is used for the structural check
against Table 1:

```{r simulate-typical}
mod_typical <- rxode2::zeroRe(mod)
# Dense grid for the typical-value profiles used in the figures.
ev_typ_dense <- bind_rows(
  lapply(c(0, 1), function(sdt) {
    tibble(
      id = sdt + 1L,
      time = c(0, seq(0.02, 24, by = 0.02)),
      amt = NA_real_, evid = 0L, cmt = "Cc",
      CONMED_SDT = sdt, arm = if (sdt == 0) "Control" else "SDT"
    ) |>
      bind_rows(tibble(
        id = sdt + 1L, time = 0, amt = tegafur_dose_ug_kg,
        evid = 1L, cmt = "depot", CONMED_SDT = sdt,
        arm = if (sdt == 0) "Control" else "SDT"
      )) |>
      arrange(time, desc(evid))
  })
)

sim_typ <- rxode2::rxSolve(
  mod_typical, events = ev_typ_dense, keep = c("arm", "CONMED_SDT")
) |>
  as.data.frame()
```

# Replicate published figures

```{r figure-1b}
# Replicates Figure 1B of Kim 2017: mean plasma tegafur and 5-FU vs time after
# oral S-1, control vs seven-day SDT pretreatment.
sim |>
  select(id, time, arm, Tegafur = Cc, `5-FU` = Cc_5fu) |>
  pivot_longer(c(Tegafur, `5-FU`), names_to = "analyte", values_to = "conc") |>
  mutate(analyte = factor(analyte, levels = c("Tegafur", "5-FU"))) |>
  group_by(analyte, arm, time) |>
  summarise(
    Q10 = quantile(conc, 0.10), Q50 = quantile(conc, 0.50),
    Q90 = quantile(conc, 0.90), .groups = "drop"
  ) |>
  filter(time > 0) |>
  ggplot(aes(time, Q50, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.2) +
  facet_wrap(~analyte, scales = "free_y") +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Plasma concentration (ng/mL)",
    colour = NULL, fill = NULL,
    title = "Figure 1B -- tegafur and 5-FU after oral S-1",
    caption = paste(
      "Replicates Figure 1B of Kim 2017 (multiple-dose pretreatment arms).",
      "Median with 10th-90th percentile band,", n_per_arm, "rats per arm."
    )
  )
```

```{r figure-2-flux}
# Illustrates the Figure 2 structure: the branched absorption site sends drug
# either to tegafur central (ka) or straight to the 5-FU precursor (Ka,Met).
sim_typ |>
  select(time, arm, Depot = depot, Precursor = precursor_5fu) |>
  pivot_longer(c(Depot, Precursor), names_to = "state", values_to = "amount") |>
  filter(time <= 12) |>
  ggplot(aes(time, amount, colour = arm)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~state, scales = "free_y") +
  labs(
    x = "Time (h)", y = "Amount (ug/kg)", colour = NULL,
    title = "Branched absorption site and the 5-FU precursor pool",
    caption = "Typical-value profiles of the states in Kim 2017 Figure 2."
  )
```

# PKNCA validation

One PKNCA block per analyte, each grouped by arm so the results line up with
Table 1's per-group columns.

```{r pknca-setup}
dose_df <- events |>
  filter(evid == 1) |>
  select(id, time, amt, arm)

make_nca <- function(conc_col) {
  # Only `!is.na()` -- filtering on time > 0 or conc > 0 would drop the
  # time-zero row PKNCA needs to anchor AUC from zero.
  d <- sim |>
    filter(!is.na(.data[[conc_col]])) |>
    transmute(id, time, arm, Cc = .data[[conc_col]])
  d <- bind_rows(
    d, d |> distinct(id, arm) |> mutate(time = 0, Cc = 0)
  ) |>
    distinct(id, arm, time, .keep_all = TRUE) |>
    arrange(id, arm, time)

  conc_obj <- PKNCA::PKNCAconc(d, Cc ~ time | arm + id)
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)
  intervals <- data.frame(
    start = 0, end = Inf,
    cmax = TRUE, tmax = TRUE, auclast = TRUE,
    aucinf.obs = TRUE, half.life = TRUE
  )
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}

nca_teg <- make_nca("Cc")
nca_5fu <- make_nca("Cc_5fu")
```

## Comparison against published NCA

Table 1 of Kim 2017 reports the multiple-dose arms, which are the arms the
population model stratifies on. `AUCinf` is the comparator for AUC: the
paper's `AUCall` stops at the last quantifiable sample, whereas the simulated
profiles are noise-free and are integrated to infinity.

```{r nca-compare-tegafur}
published_teg <- tibble::tribble(
  ~arm,      ~cmax, ~tmax, ~aucinf.obs, ~half.life,
  "Control", 6828.0, 1.2,   43748.1,     2.2,
  "SDT",     4960.0, 3.2,   47461.3,     3.3
)

cmp_teg <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_teg,
  reference = published_teg,
  by = "arm",
  units = c(
    cmax = "ng/mL", aucinf.obs = "ng*h/mL",
    tmax = "h", half.life = "h"
  ),
  tolerance_pct = 20
)

knitr::kable(
  cmp_teg,
  caption = paste(
    "Tegafur: simulated vs Kim 2017 Table 1 (multiple-dose arms).",
    "* differs from reference by >20%."
  )
)
```

```{r nca-compare-5fu}
published_5fu <- tibble::tribble(
  ~arm,      ~cmax, ~tmax, ~aucinf.obs, ~half.life,
  "Control", 172.2,  1.8,   750.2,       3.0,
  "SDT",      64.3,  2.4,   429.6,       3.1
)

cmp_5fu <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_5fu,
  reference = published_5fu,
  by = "arm",
  units = c(
    cmax = "ng/mL", aucinf.obs = "ng*h/mL",
    tmax = "h", half.life = "h"
  ),
  tolerance_pct = 20
)

knitr::kable(
  cmp_5fu,
  caption = paste(
    "5-FU: simulated vs Kim 2017 Table 1 (multiple-dose arms).",
    "* differs from reference by >20%."
  )
)
```

## Structural gate: typical-value AUC against Table 1

The typical-value prediction contains no between-subject variability, so the
comparison against the published means is a deterministic structural check
(a mis-transcribed clearance, dose or unit moves it by tens of percent) and is
asserted tightly. The per-subject cohort statistics in the section above are
not asserted on, because the extreme of a random cohort is not reproducible
across rxode2 builds.

```{r structural-gate}
trap <- function(x, y) sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)

typ <- sim_typ |>
  group_by(arm) |>
  summarise(
    teg_auc = trap(time, Cc),
    fu_auc = trap(time, Cc_5fu),
    teg_cmax = max(Cc),
    fu_cmax = max(Cc_5fu),
    .groups = "drop"
  )

gate <- typ |>
  left_join(
    published_teg |> select(arm, teg_ref = aucinf.obs, teg_cmax_ref = cmax),
    by = "arm"
  ) |>
  left_join(
    published_5fu |> select(arm, fu_ref = aucinf.obs, fu_cmax_ref = cmax),
    by = "arm"
  ) |>
  mutate(
    teg_auc_pct = 100 * (teg_auc - teg_ref) / teg_ref,
    fu_auc_pct = 100 * (fu_auc - fu_ref) / fu_ref,
    teg_cmax_pct = 100 * (teg_cmax - teg_cmax_ref) / teg_cmax_ref,
    fu_cmax_pct = 100 * (fu_cmax - fu_cmax_ref) / fu_cmax_ref
  )

gate |>
  select(arm, teg_auc_pct, fu_auc_pct, teg_cmax_pct, fu_cmax_pct) |>
  rename(
    "Arm" = arm,
    "Tegafur AUC (% diff)" = teg_auc_pct,
    "5-FU AUC (% diff)" = fu_auc_pct,
    "Tegafur Cmax (% diff)" = teg_cmax_pct,
    "5-FU Cmax (% diff)" = fu_cmax_pct
  ) |>
  knitr::kable(
    digits = 1,
    caption = "Typical-value prediction vs Kim 2017 Table 1 multiple-dose means."
  )

stopifnot(
  # Tegafur AUC is the tightest constraint in the model: it is Dose/(CL/F)
  # and is reproduced to well under one percent in both arms.
  all(abs(gate$teg_auc_pct) < 3),
  # 5-FU AUC depends on the full formation cascade; observed agreement is
  # about +1% (Control) and -3% (SDT).
  all(abs(gate$fu_auc_pct) < 10),
  # Cmax additionally depends on the distribution phase and on the paper's
  # Cmax being a mean of individual observed peaks read off a sparse grid.
  all(abs(gate$teg_cmax_pct) < 15),
  all(abs(gate$fu_cmax_pct) < 15)
)
```

## The SDT effect

The two headline effects the paper reports from the population model are
reproduced exactly, because they are ratios of tabulated parameters.

```{r sdt-effect}
ini_tab <- as.data.frame(ui$iniDf)
val <- function(nm) exp(ini_tab$est[match(nm, ini_tab$name)])

cl_ratio <- val("lcl_5fu_sdt") / val("lcl_5fu_ctl")
ka_ratio <- val("lka_sdt") / val("lka_ctl")

tibble::tibble(
  Quantity = c(
    "5-FU clearance, SDT / control",
    "Tegafur absorption rate constant, SDT / control"
  ),
  Model = c(cl_ratio, ka_ratio),
  Published = c(1.68, NA_real_)
) |>
  knitr::kable(digits = 3, caption = "Kim 2017 Section 2.3 / Discussion.")

stopifnot(
  # Kim 2017: "The clearance of 5-FU in the SDT pretreatment group was
  # 1.68-fold higher than that in the control group."
  abs(cl_ratio - 1.68) < 0.01,
  # Kim 2017: "the Ka of the SDT pretreatment group was slower".
  ka_ratio < 1
)
```

# Errata and cross-checks

## Table 1's AUC-ratio row is a molar ratio

Table 1 reports an `AUCmeta/AUCparent (%)` row whose values (2.5, 3.3, 2.7,
1.5%) are about 1.5-fold larger than the ratio of the tabulated `AUCinf`
columns. The factor is the molecular-weight ratio of tegafur (200.17 g/mol) to
5-FU (130.08 g/mol): the row is a **molar** ratio while the AUC columns are
mass-based. All four arms reconcile within rounding, so the row is internally
consistent once read on the molar scale. The model carries mass units
throughout (as the paper's equations do), so a reader comparing the model's
metabolite-to-parent AUC ratio against that row must apply the same factor.

```{r molar-ratio}
mw_ratio <- 200.17 / 130.08
tibble::tribble(
  ~arm,                     ~teg_auc, ~fu_auc, ~printed_pct,
  "Control, single dose",    55712.7,   924.9,  2.5,
  "SDT, single dose",        42945.5,   955.8,  3.3,
  "Control, multiple dose",  43748.1,   750.2,  2.7,
  "SDT, multiple dose",      47461.3,   429.6,  1.5
) |>
  mutate(
    mass_pct = 100 * fu_auc / teg_auc,
    molar_pct = mass_pct * mw_ratio
  ) |>
  select(arm, printed_pct, mass_pct, molar_pct) |>
  rename(
    "Arm" = arm, "Table 1 row (%)" = printed_pct,
    "Mass ratio (%)" = mass_pct, "Molar ratio (%)" = molar_pct
  ) |>
  knitr::kable(digits = 2, caption = "Table 1's AUC-ratio row reads as molar.")
```

```{r molar-ratio-gate}
chk <- c(55712.7, 42945.5, 43748.1, 47461.3)
met <- c(924.9, 955.8, 750.2, 429.6)
printed <- c(2.5, 3.3, 2.7, 1.5)
molar <- 100 * met / chk * mw_ratio
# Each printed value is recovered to within its own rounding step (0.05 pp)
# plus the rounding already present in the AUC columns.
stopifnot(all(abs(molar - printed) < 0.15))
```

# Assumptions and deviations

- **Between-subject variability scale.** Table 2 reports one BSV per parameter
  in a parenthesised `Population Mean (BSV)` column of bare decimals with no
  percent sign and no units, and Section 4.5 states only that "log-normal
  distribution was used to describe the between subject variability". The
  values are encoded here as **log-scale variances** (`omega^2`), which is what
  the `eta ~ value` syntax takes. The alternative reading -- that the column is
  a coefficient of variation expressed as a fraction -- would require
  `eta ~ log(1 + cv^2)` instead. The variance reading was chosen because the
  column carries no `%` header and because the smallest entry, 0.00221 for
  tegafur clearance, corresponds to a 0.22% coefficient of variation under the
  CV reading, which is not a value an estimator reports; as a variance it is a
  modest 4.7% CV. A cohort-spread falsification test was run and was not
  decisive: with the variance reading the BSV-only spread in tegafur Cmax
  (43%) exceeds the observed total spread in the control arm (33%, n = 5), but
  the two identically-treated control groups in Table 1 give Cmax CVs of 33%
  and 5.6%, so the n = 5 estimate is far too unstable to falsify either
  reading, and the model was fitted to a pooled dataset spanning separately
  conducted experiments, which inflates BSV relative to any one group.
  **This reading is flagged for confirmation**; only the stochastic cohort
  figures and the PKNCA spread depend on it, not the structural gate.
- **Residual error is not published.** Section 4.5 states that an additive plus
  proportional residual model was used for both analytes, but Table 2 reports
  no residual estimates and no supplement exists. The error structure is
  preserved with `addSd`, `propSd`, `addSd_5fu` and `propSd_5fu` all
  `fixed(0)`; no magnitudes were invented. A downstream user who wants
  realistic observation noise must supply their own.
- **Which arms entered the fit.** Section 4.5 says the oral tegafur and 5-FU
  profiles were fitted along with a historical intravenous 5-FU profile, but
  does not state whether the single-dose pretreatment groups were included
  alongside the seven-day groups. The covariate stratification has only two
  levels and the reported SDT effect matches the multiple-dose NCA result (a
  single SDT dose produced no significant change, Section 2.1), so
  `CONMED_SDT` is documented as the seven-day repeated-dose indicator and the
  population metadata records n = 10 for the oral arms.
- **`FMet` is a shadow flux, not a split of tegafur elimination.** Equation 2
  removes the full `CLTeg` from the tegafur central compartment, while
  equation 4 *adds* `CLTeg * FMet * C1,Teg` to the precursor pool. Tegafur mass
  is therefore not conserved into the metabolite, which is the usual
  bookkeeping when metabolite parameters are apparent (`/F`) and the
  parent-to-metabolite molecular-weight conversion is absorbed into the
  estimate. The equations are encoded exactly as printed.
- **No molecular-weight conversion between analytes.** Consistent with the
  point above, mass flows from the tegafur compartments into the 5-FU
  compartments without an MW factor, as the paper's equations specify.
- **Methods-versus-Table conflict on which parameters are arm-specific.**
  Section 4.5 states that `FMet` "was estimated separately for control and SDT
  pretreated rats", but Table 2 lists a single `FMet` (0.342) and instead
  splits `CL5FU/F` by arm, and Section 2.3 says explicitly that "the oral
  absorption rate of tegafur (Ka), metabolic conversion rate ... (Ka,Met), and
  the clearance of 5-FU (CL5FU) were separately estimated for control and SDT
  pretreated group". Two independent statements (Table 2 and Section 2.3) agree
  against one sentence in Section 4.5, so the model splits `ka`,
  `k_precursor_5fu_form` and `cl_5fu` and shares `fm`.
- **Non-canonical compartment pending ratification.** The 5-FU precursor pool
  (5'-hydroxytegafur) has no volume and no measured concentration, so it is not
  a `central_<metab>` state. It is named `precursor_5fu` and declared in
  `paper_specific_compartments` pending operator ratification as a canonical
  compartment; the numbered `precursor<n>` canonical denotes indirect-response
  maturation chains and is a different role. The two formation rate constants
  follow the library's existing `k_<destination>_form` grammar
  (`Urien_2005_capecitabine.R`, `NA_NA_lidocaine.R`).
- **Gimeracil is out of scope of the model.** The paper's mechanistic
  explanation runs through reduced gimeracil absorption, but gimeracil was
  analysed only noncompartmentally (Table 1) and no gimeracil structural model
  is reported, so none is encoded.
- **Observation rows name the observable, not the ODE state.** This model
  declares two endpoints (`Cc` and `Cc_5fu`), so rxode2 requires observation
  records to carry `cmt = "Cc"`; `cmt = "central"` fails with
  `'dvid'->'cmt' or 'cmt' on observation record or on a undefined compartment`.
  The static vignette pre-lint flags `cmt =` on an algebraic observable as a
  compartment-renumbering risk, which is the correct default for a
  single-endpoint model but is a false positive here. Both analytes are
  returned as columns at every output row, so one set of `cmt = "Cc"` rows
  serves both PKNCA blocks.
- **Cohort composition.** 200 virtual rats per arm at the paper's own sampling
  times; the source used n = 5 per arm. Body weight is not a covariate in the
  model (all parameters are already per-kg), so no weight distribution is
  simulated.
