---
title: "Zolpidem CNS pharmacodynamics (Kim 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Zolpidem CNS pharmacodynamics (Kim 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui_pk   <- rxode2::rxode(readModelDb("Kim_2026_zolpidem"))
ui_dsst <- rxode2::rxode(readModelDb("Kim_2026_zolpidem_dsst"))
ui_crt  <- rxode2::rxode(readModelDb("Kim_2026_zolpidem_crt"))
ui_vas  <- rxode2::rxode(readModelDb("Kim_2026_zolpidem_vas"))
cat("* Citation: ", ui_pk$reference, "\n", sep = "")
cat("* Article: <https://doi.org/10.1002/psp4.70208>\n")
cat("* Supplementary Data S1-S4: the four deposited NONMEM control streams\n")
```

Kim 2026 gave 30 healthy Korean volunteers a single 10 mg oral dose of
immediate-release zolpidem and measured three central-nervous-system
responses alongside plasma concentration: the digit symbol substitution test
(DSST, cognitive), choice reaction time (CRT, psychomotor) and a self-rated
sedation visual analog scale (VAS, subjective). The question behind the paper
is regulatory: the U.S. FDA halved the recommended starting dose for women in
2013, while the EMA and the Korean MFDS did not, and the authors ask whether
the apparent sex difference is really sex or is instead mediated by body
weight, albumin, age and baseline CNS performance.

Each PD endpoint was fitted **sequentially** to the same final PK model, as
four separate NONMEM runs. That structure is reproduced here as four model
files rather than one, because the PK parameters are *estimated* in the first
run and *fixed* in the other three -- a distinction that a single merged file
could not express:

| Model file | Deposited control stream | What is estimated |
|---|---|---|
| `Kim_2026_zolpidem` | Data S1 | The PK parameters themselves |
| `Kim_2026_zolpidem_dsst` | Data S2 | DSST response + learning effect; PK fixed |
| `Kim_2026_zolpidem_crt` | Data S3 | CRT response; PK fixed |
| `Kim_2026_zolpidem_vas` | Data S4 | VAS response + spline baseline; PK fixed |

The three PD control streams each declare a `COMP(EFFECT)` compartment, but
none of them writes a differential equation for it or ever references it: an
effect-compartment model was tested and rejected for VAS
(dOFV = -2.524, Results 3.2.4), and the declaration is a leftover.
All three final models are direct-response, so no effect compartment is
reproduced here.

**A note on time.** Methods 2.1 redefines the first Day -1 assessment
(-23 h) as *t* = 1 h, which places dosing at *t* = **24 h**. Two parts of the
model depend on that absolute clock -- the DSST learning term, and the VAS
baseline spline -- so every event table below doses at 24 h, not at 0 h.

# Population

```{r population, echo = FALSE, results = "asis"}
pop <- ui_pk$meta$population
cat("* Species: ", pop$species, "\n", sep = "")
cat("* Subjects: ", pop$n_subjects, " (", pop$sex_female_pct, "% female)\n", sep = "")
cat("* Age: ", pop$age_range, " (median ", pop$age_median, ")\n", sep = "")
cat("* Weight: ", pop$weight_range, " (median ", pop$weight_median, ")\n", sep = "")
cat("* Disease state: ", pop$disease_state, "\n", sep = "")
cat("* Dose: ", pop$dose_range, "\n", sep = "")
cat("* Region: ", pop$regions, "\n", sep = "")
```

Kim 2026 Table 1 gives baseline characteristics as median (min to max), both
overall and by sex. Weight, albumin, ALT, AST, total bilirubin and serum
creatinine all differed significantly between males and females, which is
exactly the collinearity that motivates the paper.

```{r population-table, echo = FALSE}
tibble::tribble(
  ~Variable,                    ~Total,                  ~Male,                  ~Female,
  "Age (years)",                "29.5 (20 to 44)",       "31 (20 to 42)",        "27 (24 to 44)",
  "Weight (kg)",                "60.05 (50.0 to 83.0)",  "75.4 (55.1 to 83.0)",  "54.9 (50.0 to 61.1)",
  "Albumin (g/dL)",             "4.4 (3.9 to 5.0)",      "4.5 (4.1 to 5.0)",     "4.2 (3.9 to 4.8)",
  "DSST at 1 h Day -1 (score)", "72.5 (46 to 104)",      "63 (46 to 83)",        "74 (60 to 104)",
  "CRT at 1 h Day -1 (msec)",   "438.5 (372 to 709)",    "438 (402 to 709)",     "453 (372 to 521)",
  "VAS at 1 h Day -1 (mm)",     "30.5 (0 to 97)",        "24 (0 to 97)",         "48 (0 to 72)"
) |>
  knitr::kable(caption = "Kim 2026 Table 1 (subset): baseline characteristics, median (min to max).")
```

# Source trace

Every value in the four model files is traceable to Kim 2026 Table 2 or
Table 3 and, independently, to the corresponding `$THETA` / `$OMEGA` entry of
the deposited control stream. The two sources agree on every parameter except
one, noted below.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Quantity, ~`Model parameter`, ~`Source location`,
  "Transit absorption input", "`transit(nn, mtt, fdepot)`", "Data S1 `$DES` (Savic closed form); Figure 1",
  "One-compartment disposition", "`d/dt(central)`", "Data S1 `$DES`; Results 3.2.1",
  "Ka 11.7 1/h", "`lka`", "Table 2; Data S1 `$THETA(3)`",
  "CL/F 18.0 L/h", "`lcl`", "Table 2; Data S1 `$THETA(1)`",
  "Vc/F 64.0 L", "`lvc`", "Table 2; Data S1 `$THETA(2)`",
  "MTT 0.25 h", "`lmtt`", "Table 2; Data S1 `$THETA(4)`",
  "NN 19.4", "`lnn`", "Table 2; Data S1 `$THETA(5)`",
  "BIO 1 (fixed)", "`lfdepot`", "Table 2; Data S1 `$THETA(6)` FIX",
  "PK proportional error 18.9%", "`propSd`", "Table 2; Data S1 `$THETA(7)`",
  "DSST: IC50 205 ug/L", "`lic50`", "Table 3 DSST block; Data S2 `$THETA(9)`",
  "DSST: HILL 2.54", "`lhill`", "Table 3 DSST block; Data S2 `$THETA(10)`",
  "DSST: BASE 68.2 score", "`lrbase`", "Table 3 DSST block; Data S2 `$THETA(11)`",
  "DSST: MAXlearn 14.4 score", "`lemax_learn`", "Table 3 DSST block; Data S2 `$THETA(12)`",
  "DSST: LRPAR 4.3 h", "`lt50_learn`", "Table 3 DSST block; Data S2 `$THETA(13)`",
  "DSST: weight on IC50, -1.57", "`e_wt_ic50`", "Table 3 row; Data S2 `$THETA(18)` (**not** footnote a -- see Errata)",
  "DSST: albumin on IC50, 0.541", "`e_alb_ic50`", "Table 3 row and footnote a; Data S2 `$THETA(17)`",
  "DSST: age on BASE, -0.0143", "`e_age_rbase`", "Table 3 row and footnote b; Data S2 `$THETA(16)`",
  "CRT: EC50 282 ug/L", "`lec50`", "Table 3 CRT block; Data S3 `$THETA(9)`",
  "CRT: HILL 6", "`lhill`", "Table 3 CRT block; Data S3 `$THETA(12)`",
  "CRT: BASE 437 msec", "`lrbase`", "Table 3 CRT block; Data S3 `$THETA(10)`",
  "CRT: MAXdrug 5.6 (fixed)", "`lemax`", "Table 3 CRT block; Data S3 `$THETA(11)` FIX; Results 3.2.3",
  "CRT: weight on EC50, -0.0193", "`e_wt_ec50`", "Table 3 row and footnote c; Data S3 `$THETA(18)`",
  "CRT: albumin on EC50, 3.89", "`e_alb_ec50`", "Table 3 row and footnote c; Data S3 `$THETA(16)`",
  "CRT: CRTB1 on EC50, -1.28", "`e_crt_bl_ec50`", "Table 3 row and footnote c; Data S3 `$THETA(17)`",
  "CRT: albumin on HILL, -5.26", "`e_alb_hill`", "Table 3 row and footnote d; Data S3 `$THETA(19)`",
  "CRT: CRTB1 on HILL, 0.00637", "`e_crt_bl_hill`", "Table 3 row and footnote d; Data S3 `$THETA(20)`",
  "CRT: age on BASE, 0.0112", "`e_age_rbase`", "Table 3 row and footnote e; Data S3 `$THETA(15)`",
  "VAS: EC50 146 ug/L", "`lec50`", "Table 3 VAS block; Data S4 `$THETA(9)`",
  "VAS: HILL 1.92", "`lhill`", "Table 3 VAS block; Data S4 `$THETA(10)`",
  "VAS: knots H1-H4", "`logitrbase_t1`..`_t4`", "Table 3 VAS block and Results 3.2.4; Data S4 `$THETA(11)`-`$THETA(14)`",
  "VAS: logit transform of knots", "`expit(logitrbase_ti + eta)`", "Methods 2.3; Data S4 `$PK`",
  "VAS: piecewise-linear spline", "`vasbase`", "Results 3.2.4 displayed equations; Data S4 `$ERROR`",
  "VAS: headroom-bounded drug effect", "`(100 - vasbase) * ...`", "Results 3.2.4; Data S4 `$ERROR`"
) |>
  knitr::kable(caption = "Source trace for the model equations and parameters.")
```

## Reconciling the reported %CV against the encoded variances

Kim 2026 reports most inter-individual variabilities as %CV, with the note
under Table 3 giving the transform used. The control-stream `$OMEGA` entries
are variances. This block checks every reported %CV against the variance
actually encoded in the model files -- an enumerating check, so a
mis-transcribed `omega` fails here rather than passing silently.

```{r cv-check}
omega_of <- function(ui, eta) {
  df <- ui$iniDf
  df$est[!is.na(df$neta1) & df$neta1 == df$neta2 & df$name == eta]
}
pct_cv <- function(omega2) 100 * sqrt(exp(omega2) - 1)

cv_chk <- tibble::tribble(
  ~Model,  ~Parameter,   ~eta,           ~ui,      ~reported_cv,
  "PK",    "Ka",         "etalka",       "pk",     330.8,
  "PK",    "MTT",        "etalmtt",      "pk",      37.7,
  "PK",    "CL/F",       "etalcl",       "pk",      24.1,
  "PK",    "BIO",        "etalfdepot",   "pk",      25.9,
  "DSST",  "IC50",       "etalic50",     "dsst",    22.3,
  "DSST",  "BASE",       "etalrbase",    "dsst",    13.1,
  "CRT",   "EC50",       "etalec50",     "crt",     21.1,
  "CRT",   "BASE",       "etalrbase",    "crt",      9.4,
  "VAS",   "EC50",       "etalec50",     "vas",     53.9
) |>
  rowwise() |>
  mutate(
    omega2      = omega_of(get(paste0("ui_", ui)), eta),
    computed_cv = pct_cv(omega2)
  ) |>
  ungroup() |>
  mutate(abs_diff = abs(computed_cv - reported_cv))

# Every reported %CV is printed to 3 significant figures, so agreement must be
# within half a unit in the last printed place.
stopifnot(all(cv_chk$abs_diff < 0.05))

cv_chk |>
  select(Model, Parameter, omega2, computed_cv, reported_cv) |>
  mutate(across(c(omega2, computed_cv), ~ signif(.x, 4))) |>
  rename(
    "Encoded variance" = omega2,
    "%CV computed"     = computed_cv,
    "%CV reported"     = reported_cv
  ) |>
  knitr::kable(caption = "Every %CV in Kim 2026 Tables 2 and 3 reproduces from the encoded variance via 100 * sqrt(exp(omega^2) - 1).")
```

The four VAS baseline knots are the exception the paper flags: Table 3
footnote f states their IIV is "expressed as omega^2 instead of %CV",
so those four are read straight off the table with no transform.

```{r vas-omega-check}
vas_knot_omegas <- vapply(
  paste0("etalogitrbase_t", 1:4), omega_of, numeric(1), ui = ui_vas
)
# Table 3: 1.03, 0.653, 1.7, 0.699. The control stream gives 1.69 for the third,
# which is the unrounded value behind the table's 1.7.
stopifnot(all(abs(vas_knot_omegas - c(1.03, 0.653, 1.69, 0.699)) < 1e-12))
vas_knot_omegas
```

# Virtual cohort

The cohort is rebuilt to Kim 2026 Table 1: 15 males and 15 females, with the
sex-specific medians and ranges reproduced for the covariates the models
actually use. Weight, albumin, age and baseline CRT are drawn from truncated
normals calibrated so that the simulated median lands on the published median
and the draw stays inside the published min-max. That distributional choice is
an assumption -- Table 1 reports only medians and ranges, not shapes.

Note that `ALB` is carried in the canonical register unit **g/L**, while
Kim 2026 reports g/dL; the model files apply the register-mandated inline
conversion, so the cohort below stores 44 g/L where the paper prints 4.4 g/dL.

```{r cohort}
set.seed(20260901)

rtrunc_median <- function(n, med, lo, hi) {
  # SD chosen so the published min-max spans roughly +/- 2 SD.
  x <- stats::rnorm(n, mean = med, sd = (hi - lo) / 4)
  pmin(pmax(x, lo), hi)
}

n_per_sex <- 15
cohort <- bind_rows(
  tibble::tibble(
    SEXF   = 0,
    WT     = rtrunc_median(n_per_sex, 75.4, 55.1, 83.0),
    ALB    = rtrunc_median(n_per_sex,  4.5,  4.1,  5.0) * 10, # g/dL -> canonical g/L
    AGE    = rtrunc_median(n_per_sex, 31,   20,   42),
    CRT_BL = rtrunc_median(n_per_sex, 438,  402,  709)
  ),
  tibble::tibble(
    SEXF   = 1,
    WT     = rtrunc_median(n_per_sex, 54.9, 50.0, 61.1),
    ALB    = rtrunc_median(n_per_sex,  4.2,  3.9,  4.8) * 10,
    AGE    = rtrunc_median(n_per_sex, 27,   24,   44),
    CRT_BL = rtrunc_median(n_per_sex, 453,  372,  521)
  )
) |>
  mutate(id = row_number(), sex = ifelse(SEXF == 1, "Female", "Male"))

cohort |>
  group_by(sex) |>
  summarise(
    n = n(),
    `Weight (kg)`      = sprintf("%.1f (%.1f-%.1f)", median(WT), min(WT), max(WT)),
    `Albumin (g/dL)`   = sprintf("%.2f (%.2f-%.2f)", median(ALB) / 10, min(ALB) / 10, max(ALB) / 10),
    `Age (years)`      = sprintf("%.0f (%.0f-%.0f)", median(AGE), min(AGE), max(AGE)),
    `CRT_BL (msec)`    = sprintf("%.0f (%.0f-%.0f)", median(CRT_BL), min(CRT_BL), max(CRT_BL)),
    .groups = "drop"
  ) |>
  rename(Sex = sex) |>
  knitr::kable(caption = "Simulated cohort, median (min-max), to be read against Kim 2026 Table 1.")
```

# Pharmacokinetics

## Simulation

The PK model carries no absolute-time term, so the PK simulation doses at
*t* = 0 and observes over the paper's Day 1 sampling window (0-12 h). The PD
simulations further down dose at *t* = 24 h, as the PD models require.

```{r pk-sim}
pk_grid <- sort(unique(c(seq(0, 12, by = 0.05),
                         c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 12))))

pk_events <- bind_rows(
  cohort |> transmute(id, time = 0, cmt = "depot", amt = 10, evid = 1L),
  tidyr::expand_grid(id = cohort$id, time = pk_grid) |>
    mutate(cmt = "central", amt = 0, evid = 0L)
) |>
  left_join(cohort |> select(id, WT, ALB, AGE, CRT_BL, SEXF, sex), by = "id") |>
  arrange(id, time, desc(evid)) |>
  as.data.frame()

rxode2::rxSetSeed(20260901)
pk_sim <- rxode2::rxSolve(ui_pk, pk_events, returnType = "data.frame") |>
  left_join(cohort |> select(id, sex), by = "id")

# rxode2's transit() has been observed to evaluate silently to zero through an
# rxUi handle on some builds. Assert that it did not, so this vignette can never
# ship an all-zero profile that merely looks like a flat plot.
stopifnot(max(pk_sim$Cc) > 1)
```

```{r pk-typical}
# Typical-value profile (all etas zeroed) at the cohort-median covariates.
typ_events <- data.frame(
  id = 1L,
  time = c(0, pk_grid),
  cmt = c("depot", rep("central", length(pk_grid))),
  amt = c(10, rep(0, length(pk_grid))),
  evid = c(1L, rep(0L, length(pk_grid)))
)
pk_typ <- rxode2::rxSolve(rxode2::zeroRe(ui_pk), typ_events, returnType = "data.frame")
```

## Figure: plasma zolpidem concentration-time profile

Replicates the observation window of Kim 2026 Figure 2A (prediction-corrected
VPC for plasma zolpidem) and Figure S2A.

```{r fig-pk, fig.cap = "Simulated plasma zolpidem concentrations after a single 10 mg oral dose. Ribbon = 5th-95th percentile of the simulated cohort, solid line = median, dashed line = typical-value profile. Replicates the layout of Kim 2026 Figure 2A."}
# rxSolve returns one row per observation record, so no dose rows to filter out
# (and no `evid` column in the output).
pk_band <- pk_sim |>
  group_by(time) |>
  summarise(
    lo  = quantile(Cc, 0.05),
    mid = median(Cc),
    hi  = quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot(pk_band, aes(time)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, fill = "steelblue") +
  geom_line(aes(y = mid), colour = "steelblue", linewidth = 1) +
  geom_line(data = pk_typ, aes(time, Cc), linetype = "dashed") +
  labs(x = "Time after dose (h)", y = "Plasma zolpidem (ug/L)") +
  theme_bw()
```

## PKNCA validation

```{r pknca}
nca_conc <- pk_sim |>
  filter(!is.na(Cc)) |>
  transmute(id, time, Cc, treatment = "Zolpidem 10 mg single oral dose")

nca_dose <- pk_events |>
  filter(evid == 1) |>
  transmute(id, time, amt, treatment = "Zolpidem 10 mg single oral dose")

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ time | treatment + id,
                             concu = "ug/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(nca_dose, amt ~ time | treatment + id,
                             doseu = "mg")

intervals <- data.frame(
  start       = 0,
  end         = Inf,
  cmax        = TRUE,
  tmax        = TRUE,
  auclast     = TRUE,
  aucinf.obs  = TRUE,
  half.life   = TRUE,
  lambda.z    = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_ind <- as.data.frame(nca_res)
```

### Structural gate: AUC0-inf must equal Dose x F / CL, per subject

This is a self-consistency check rather than a comparison against a published
number, and the two sides use the *same* drawn parameters, so a tight bound is
the right one: the only difference is numerical (integration tolerance plus
PKNCA's terminal extrapolation from the 12 h profile). If the transit chain
failed to deliver the whole dose, or the concentration scaling were wrong, this
fails immediately.

```{r auc-gate}
auc_chk <- nca_ind |>
  filter(PPTESTCD == "aucinf.obs") |>
  transmute(id = as.integer(as.character(id)), aucinf = PPORRES) |>
  left_join(
    pk_sim |> group_by(id) |> summarise(cl = first(cl), fdepot = first(fdepot), .groups = "drop"),
    by = "id"
  ) |>
  mutate(
    auc_closed_form = 10 * fdepot / cl * 1000, # mg / (L/h) -> ug*h/L
    pct_diff = 100 * (aucinf - auc_closed_form) / auc_closed_form
  )

stopifnot(
  abs(median(auc_chk$pct_diff)) < 1,
  quantile(abs(auc_chk$pct_diff), 0.9) < 2
)

sprintf("AUC0-inf vs Dose*F/CL: median %.3f%%, 90th pctile |diff| %.3f%%",
        median(auc_chk$pct_diff), quantile(abs(auc_chk$pct_diff), 0.9))
```

### Comparison against the published non-compartmental analysis

Kim 2026 does not tabulate its own NCA; the Introduction quotes the NCA of the
source clinical study (reference 11, the same 30 subjects and the same 10 mg
dose) for T~max~ and half-life, and those are the published values compared
against here.

```{r nca-comparison}
nca_simulated <- nca_ind |>
  filter(PPTESTCD %in% c("cmax", "tmax", "half.life", "aucinf.obs")) |>
  group_by(PPTESTCD) |>
  summarise(PPORRES = median(PPORRES), .groups = "drop") |>
  as.data.frame()

# Tmax reference is the midpoint of the published 0.5-0.75 h median range and
# half-life the midpoint of the published 2.9-3.1 h mean range, both quoted in
# Kim 2026 Introduction from the source clinical study (reference 11). Cmax and
# AUC0-inf were not reported there, so only the two published parameters can be
# compared.
nca_reference <- data.frame(
  PPTESTCD = c("tmax", "half.life"),
  PPORRES  = c(0.625, 3.0)
)

nlmixr2lib::ncaComparisonTable(
  simulated = nca_simulated,
  reference = nca_reference,
  units = c(cmax = "ug/L", tmax = "h", half.life = "h", aucinf.obs = "ug*h/L")
) |>
  knitr::kable(
    caption = paste(
      "Simulated median NCA parameters against the published values.",
      "Tmax reference is the midpoint of the reported 0.5-0.75 h median range;",
      "half-life reference is the midpoint of the reported 2.9-3.1 h mean range",
      "(Kim 2026 Introduction, citing the source clinical study)."
    )
  )
```

For completeness, the simulated median C~max~ and AUC~0-inf~, neither of which
the paper reports:

```{r nca-unreported, echo = FALSE}
nca_simulated |>
  filter(PPTESTCD %in% c("cmax", "aucinf.obs")) |>
  mutate(
    Parameter = c("AUC0-inf (ug*h/L)", "Cmax (ug/L)")[match(PPTESTCD, c("aucinf.obs", "cmax"))],
    Simulated = signif(PPORRES, 4)
  ) |>
  select(Parameter, Simulated) |>
  knitr::kable(caption = "Simulated median exposure metrics not reported by Kim 2026.")
```

The simulated median T~max~ falls inside the published 0.5-0.75 h window. The
simulated half-life is shorter than the published NCA mean: the model's
disposition half-life is fixed by ln(2) x V~c~/F / (CL/F) = ln(2) x 64 / 18 =
`r sprintf("%.2f", log(2) * 64 / 18)` h, whereas a non-compartmental
lambda~z~ fitted to the 8-12 h tail of the observed data recovers a
slightly shallower slope. This is a property of the published one-compartment
structural choice, not a transcription error, and it is reported rather than
tuned away.

```{r halflife-identity}
# The model's disposition half-life is an exact identity, so gate it exactly.
theta <- ui_pk$theta
stopifnot(abs(log(2) * exp(theta[["lvc"]]) / exp(theta[["lcl"]]) - log(2) * 64 / 18) < 1e-9)
```

# Pharmacodynamics

## Simulation

All three PD models share the dose at *t* = 24 h. The observation grid covers
the Day -1 baseline period (0-24 h) and the Day 1 post-dose period
(24-36 h), and includes the four spline knot times on both days.

```{r pd-sim}
pd_grid <- sort(unique(c(
  seq(0, 36, by = 0.25),
  c(1, 2, 4, 12),           # Day -1 knots
  24 + c(0.5, 1, 1.5, 2, 3, 4, 6, 8, 12) # Day 1 PD sampling schedule
)))

pd_events <- bind_rows(
  cohort |> transmute(id, time = 24, cmt = "depot", amt = 10, evid = 1L, dvid = NA_integer_),
  tidyr::expand_grid(id = cohort$id, time = pd_grid) |>
    mutate(cmt = "central", amt = 0, evid = 0L, dvid = 1L)
) |>
  left_join(cohort |> select(id, WT, ALB, AGE, CRT_BL, SEXF), by = "id") |>
  arrange(id, time, desc(evid)) |>
  as.data.frame()

solve_pd <- function(ui) {
  rxode2::rxSetSeed(20260901)
  rxode2::rxSolve(ui, pd_events, returnType = "data.frame") |>
    left_join(cohort |> select(id, sex), by = "id")
}

sim_dsst <- solve_pd(ui_dsst)
sim_crt  <- solve_pd(ui_crt)
sim_vas  <- solve_pd(ui_vas)

stopifnot(max(sim_dsst$Cc) > 1, max(sim_crt$Cc) > 1, max(sim_vas$Cc) > 1)
```

A typical-value (etas zeroed) solve at the reference covariates -- weight
60.05 kg, albumin 4.4 g/dL, age 29.5 years, baseline CRT 438.5 msec, all the
Table 1 medians -- gives the reference subject the paper describes.

```{r pd-typical}
ref_cov <- list(WT = 60.05, ALB = 44, AGE = 29.5, CRT_BL = 438.5)

ref_events <- data.frame(
  id   = 1L,
  time = c(24, pd_grid),
  cmt  = c("depot", rep("central", length(pd_grid))),
  amt  = c(10, rep(0, length(pd_grid))),
  evid = c(1L, rep(0L, length(pd_grid))),
  dvid = c(NA_integer_, rep(1L, length(pd_grid)))
)
for (nm in names(ref_cov)) ref_events[[nm]] <- ref_cov[[nm]]
ref_events <- ref_events[order(ref_events$time, -ref_events$evid), ]

typ_dsst <- rxode2::rxSolve(rxode2::zeroRe(ui_dsst), ref_events, returnType = "data.frame")
typ_crt  <- rxode2::rxSolve(rxode2::zeroRe(ui_crt),  ref_events, returnType = "data.frame")
typ_vas  <- rxode2::rxSolve(rxode2::zeroRe(ui_vas),  ref_events, returnType = "data.frame")

# At the reference covariates every covariate multiplier must collapse to 1, so
# each typical parameter must equal its printed Table 3 value exactly.
stopifnot(
  abs(typ_dsst$ic50[1]  - 205) < 1e-9,
  abs(typ_dsst$hill[1]  - 2.54) < 1e-9,
  abs(typ_dsst$rbase[1] -  68.2) < 1e-9,
  abs(typ_crt$ec50[1]   - 282) < 1e-9,
  abs(typ_crt$hill[1]   -   6) < 1e-9,
  abs(typ_crt$rbase[1]  - 437) < 1e-9,
  abs(typ_crt$emax[1]   -   5.6) < 1e-9,
  abs(typ_vas$ec50[1]   - 146) < 1e-9,
  abs(typ_vas$hill[1]   -   1.92) < 1e-9
)
```

## Gate: the potency rank order reported in the Discussion

> "the estimated potency (EC~50~ or IC~50~) and Hill coefficient were
> 205 ug/L and 2.54 for DSST, 282 ug/L and 6 for CRT,
> and 146 ug/L and 1.92 for VAS ... the rank order of potency
> (VAS > DSST > CRT)"

```{r potency-rank}
potency <- c(VAS = typ_vas$ec50[1], DSST = typ_dsst$ic50[1], CRT = typ_crt$ec50[1])
stopifnot(potency[["VAS"]] < potency[["DSST"]], potency[["DSST"]] < potency[["CRT"]])
potency
```

## Gate: the DSST learning effect at the time of dosing

> "The estimated time to half-maximal learning effect was 4.3 h, indicating
> that approximately 85% of the maximal learning effect would be achieved
> within 24 h"

This is the check that pins the learning term to **absolute study time**
rather than time after dose. If the term ran on time after dose it would be
exactly zero at the moment of dosing, not 85% complete.

```{r learning-gate}
at24 <- typ_dsst[typ_dsst$time == 24, ][1, ]
learn_fraction <- (at24$elearn - at24$rbase) / at24$emax_learn
stopifnot(abs(learn_fraction - 0.85) < 0.01)
sprintf("Learning effect accrued at the time of dosing (t = 24 h): %.1f%% of MAXlearn",
        100 * learn_fraction)
```

## Gate: the baseline VAS spline knots

> "the typical baseline VAS values at 1, 2, 4, and 12 h were 30.3, 36.2, 19.8,
> and 12.7 mm, respectively"

The four published baselines are recovered by solving the model over the
Day -1 window and reading `vasbase` at the knot times -- so this exercises the
logit parameterisation, the back-transform and the spline together, not just
the stored numbers.

```{r vas-knot-gate}
knot_times <- c(1, 2, 4, 12)
vasbase_knots <- typ_vas$vasbase[match(knot_times, typ_vas$time)]
stopifnot(all(abs(vasbase_knots - c(30.3, 36.2, 19.8, 12.7)) < 0.05))

tibble::tibble(
  `Knot (h, Day -1)` = knot_times,
  `Simulated (mm)`   = round(vasbase_knots, 2),
  `Kim 2026 (mm)`    = c(30.3, 36.2, 19.8, 12.7)
) |>
  knitr::kable(caption = "Baseline sedation VAS spline knots against Kim 2026 Results 3.2.4.")
```

## Gate: the covariate forest plots (Kim 2026 Figure 3, Results 3.4)

Results 3.4 quotes point estimates for the covariate effects at the 10th and
90th percentiles of the observed covariate distributions relative to the
reference subject. These are the strongest available checks on the covariate
equations, because the authors computed them from the same coefficients. They
are evaluated here from the **solved model's own** covariate multiplier
columns (`cov_ic50`, `cov_ec50`, `cov_hill`), not from re-typed formulas.

Two things keep this from being an exact identity, so the tolerance is 5%
relative rather than the printed precision. First, the Figure 3 dots are the
**median over 1000 SIR replicates** of the predicted relative change (Methods
2.4 and the Figure 3 caption), not the ratio evaluated at the final point
estimate; for the power terms that median is shifted by the skew of the
sampled exponent. Second, the covariate percentiles themselves are quoted to
two or three significant figures. A 5% band still discriminates decisively:
the mis-typeset linear reading of the weight effect on DSST IC~50~ would give
1.30 against a published 0.66, a 97% error.

```{r forest-gate}
cov_multiplier <- function(ui, column, changes) {
  cov <- utils::modifyList(ref_cov, changes)
  ev <- data.frame(id = 1L, time = c(24, 25), cmt = c("depot", "central"),
                   amt = c(10, 0), evid = c(1L, 0L), dvid = c(NA_integer_, 1L))
  for (nm in names(cov)) ev[[nm]] <- cov[[nm]]
  rxode2::rxSolve(rxode2::zeroRe(ui), ev, returnType = "data.frame")[[column]][1]
}

forest <- tibble::tribble(
  ~Parameter,       ~`Covariate value`,        ~ui,       ~column,     ~changes,             ~published,
  "DSST IC50",      "Weight 78.0 kg (90th)",   "dsst",    "cov_ic50",  list(WT = 78.0),      0.66,
  "CRT EC50",       "Weight 78.0 kg (90th)",   "crt",     "cov_ec50",  list(WT = 78.0),      0.65,
  "CRT HILL",       "Albumin 4.0 g/dL (10th)", "crt",     "cov_hill",  list(ALB = 40),       1.6,
  "CRT HILL",       "Albumin 4.8 g/dL (90th)", "crt",     "cov_hill",  list(ALB = 48),       0.65,
  "CRT HILL",       "CRT_BL 504 msec (90th)",  "crt",     "cov_hill",  list(CRT_BL = 504),   1.4
) |>
  rowwise() |>
  mutate(simulated = cov_multiplier(get(paste0("ui_", ui)), column, changes)) |>
  ungroup() |>
  mutate(pct_diff = 100 * (simulated - published) / published)

stopifnot(max(abs(forest$pct_diff)) < 5)

forest |>
  select(Parameter, `Covariate value`, simulated, published, pct_diff) |>
  mutate(simulated = round(simulated, 3), pct_diff = round(pct_diff, 2)) |>
  rename(
    "Simulated ratio vs reference" = simulated,
    "Kim 2026 Results 3.4"         = published,
    "% diff"                       = pct_diff
  ) |>
  knitr::kable(
    caption = paste(
      "Covariate effect ratios relative to the reference subject, against the",
      "point estimates quoted in Kim 2026 Results 3.4 (Figure 3 forest plots).",
      "Each simulated value is read from the solved model's own covariate",
      "multiplier column."
    )
  )
```

### Hard gate: power versus linear weight effect on DSST IC~50~

Kim 2026 Table 3 footnote a prints the weight term without its exponent, which
reads as a linear multiplier. The two candidate readings are not close, so this
gate is exact rather than tolerance-based: the encoded model must land on the
published 0.66 and must be nowhere near the 1.30 the footnote's literal
reading would produce.

```{r weight-form-gate}
ratio_encoded <- cov_multiplier(ui_dsst, "cov_ic50", list(WT = 78.0))
ratio_if_linear <- 78.0 / 60.05

stopifnot(
  # The encoded (power) form is within 1% of the published 0.66 ...
  abs(ratio_encoded - 0.66) / 0.66 < 0.01,
  # ... and the footnote's literal linear reading is off by more than 90%.
  abs(ratio_if_linear - 0.66) / 0.66 > 0.9,
  # The exponent itself is the value printed in the Table 3 row.
  abs(ui_dsst$theta[["e_wt_ic50"]] - (-1.57)) < 1e-12
)

sprintf("Weight effect on DSST IC50 at 78.0 kg: encoded power form %.4f vs published 0.66; literal linear reading of footnote a would give %.4f.",
        ratio_encoded, ratio_if_linear)
```

## Figure: PD time courses

Replicates the layout of Kim 2026 Figure 2B-D (prediction-corrected VPCs for
DSST, CRT and VAS) and Figure S14 (model-predicted profiles at the reference
age of 29.5 years). Time is plotted on the study clock, with dosing at 24 h.

```{r fig-pd, fig.height = 8, fig.cap = "Simulated PD time courses. Ribbon = 5th-95th percentile of the simulated cohort, solid line = median, dashed line = the reference-subject typical-value profile. The vertical line marks dosing at t = 24 h. Replicates the layout of Kim 2026 Figure 2B-D."}
band_of <- function(sim, var) {
  sim |>
    filter(!is.na(.data[[var]])) |>
    group_by(time) |>
    summarise(
      lo = quantile(.data[[var]], 0.05),
      mid = median(.data[[var]]),
      hi = quantile(.data[[var]], 0.95),
      .groups = "drop"
    ) |>
    mutate(endpoint = var)
}

typ_of <- function(typ, var) {
  typ |>
    filter(!is.na(.data[[var]])) |>
    transmute(time, value = .data[[var]], endpoint = var)
}

pd_bands <- bind_rows(
  band_of(sim_dsst, "DSST"), band_of(sim_crt, "CRT"), band_of(sim_vas, "VAS")
) |>
  mutate(endpoint = factor(endpoint, levels = c("DSST", "CRT", "VAS"),
                           labels = c("DSST (score)", "CRT (msec)", "Sedation VAS (mm)")))

pd_typ <- bind_rows(
  typ_of(typ_dsst, "DSST"), typ_of(typ_crt, "CRT"), typ_of(typ_vas, "VAS")
) |>
  mutate(endpoint = factor(endpoint, levels = c("DSST", "CRT", "VAS"),
                           labels = c("DSST (score)", "CRT (msec)", "Sedation VAS (mm)")))

ggplot(pd_bands, aes(time)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, fill = "darkorange") +
  geom_line(aes(y = mid), colour = "darkorange", linewidth = 1) +
  geom_line(data = pd_typ, aes(time, value), linetype = "dashed") +
  geom_vline(xintercept = 24, colour = "grey40", linetype = "dotted") +
  facet_wrap(~endpoint, ncol = 1, scales = "free_y") +
  labs(x = "Study time (h); dose at 24 h", y = NULL) +
  theme_bw()
```

## Gate: return to baseline by 12 h after dosing

> "At the reference age of 29.5 years, both DSST and CRT responses returned to
> baseline levels approximately 12 h after administration of zolpidem 10 mg
> (Figure S14)."

For DSST, "baseline" at 12 h post-dose is the learning-adjusted baseline
`elearn`, not the constant `BASE`; for CRT it is `rbase`. Both are read off the
typical-value solve.

```{r recovery-gate}
t_end <- 36 # 12 h after the 24 h dose
d36 <- typ_dsst[typ_dsst$time == t_end, ][1, ]
c36 <- typ_crt[typ_crt$time == t_end, ][1, ]

dsst_recovery <- d36$DSST / d36$elearn
crt_recovery  <- c36$CRT  / c36$rbase

# Both endpoints must be within 1% of their own baseline at 12 h post-dose.
stopifnot(dsst_recovery > 0.99, crt_recovery < 1.01)

sprintf("At 12 h post-dose: DSST is %.2f%% of its learning-adjusted baseline; CRT is %.2f%% of its baseline.",
        100 * dsst_recovery, 100 * crt_recovery)
```

## Figure: combined covariate effects on potency

Replicates Kim 2026 Figure 4A -- typical DSST IC~50~ as a joint function of
weight and albumin, over the observed ranges of both.

```{r fig-heatmap, fig.height = 4, fig.cap = "Typical DSST IC50 across the observed weight and albumin ranges, from the fixed effects of the final PK-DSST model. Replicates Kim 2026 Figure 4A."}
grid_cov <- tidyr::expand_grid(
  WT  = seq(50, 83, length.out = 34),
  ALB = seq(3.9, 5.0, length.out = 23) * 10
) |>
  mutate(
    ic50 = 205 *
      (WT / 60.05)^ui_dsst$theta[["e_wt_ic50"]] *
      (1 + ui_dsst$theta[["e_alb_ic50"]] * (ALB * 0.1 - 4.4))
  )

# Cross-check one grid point against the solved model so the plotted surface is
# not an independently re-typed formula.
chk_point <- grid_cov[which.min(abs(grid_cov$WT - 78) + abs(grid_cov$ALB - 44)), ]
stopifnot(abs(
  chk_point$ic50 -
    205 * cov_multiplier(ui_dsst, "cov_ic50", list(WT = chk_point$WT, ALB = chk_point$ALB))
) < 1e-8)

ggplot(grid_cov, aes(WT, ALB / 10)) +
  geom_raster(aes(fill = ic50)) +
  geom_contour(aes(z = ic50), colour = "white", linewidth = 0.3) +
  scale_fill_viridis_c(name = "IC50 (ug/L)") +
  labs(x = "Weight (kg)", y = "Albumin (g/dL)") +
  theme_bw()
```

# Assumptions and deviations

## Errata and source conflicts

1. **Kim 2026 Table 3 footnote a drops an exponent.** The footnote prints
   `IC50,DSST (ug/L) = 205 x (Weight/60.05) x (1 + 0.541 x (Albumin - 4.4))`,
   with weight entering *linearly*. The deposited control stream Data S2 gives
   the true form, `IC50WEI = ((WEI/60.05)**THETA(18))` with
   `THETA(18) = -1.57`, i.e. a *power* term; the orphaned "-1.57" that trails
   the Table 3 abbreviations block in the published PDF is the dropped
   superscript. Table 3's own "Effect of weight on IC~50~" row also reports
   -1.57. The power form is what the model file encodes, and the forest-plot
   gate above settles it independently: Results 3.4 quotes 0.66 [0.59-0.74] at
   the 90th weight percentile, and `(78.0/60.05)^-1.57` = 0.663, whereas the
   linear form printed in the footnote would give 1.30.

2. **H~3~ differs by one unit in the third digit.** Kim 2026 Table 3 reports
   the third VAS baseline knot as 0.198 (SIR median 0.198) and Results 3.2.4
   states the corresponding baseline as 19.8 mm; the Data S4 `$THETA(13)`
   entry is 0.199. The model file uses the paper's 0.198, which is stated
   three times independently against the control stream's once. The difference
   is 0.1 mm on a 100 mm scale.

3. **Near-zero fixed residual-error components are retained.** Each control
   stream builds a combined error, `W = SQRT(prop^2 * IPRED^2 + add^2)`, and
   fixes the unused half at 1e-5 as a numerical guard. Kim 2026 Tables 2 and 3
   report only the estimated half. The model files encode both, wrapping the
   1e-5 term in `fixed()`, so that the encoded error model matches the
   deposited `$ERROR` block exactly.

## Assumptions

1. **Covariate distribution shapes.** Kim 2026 Table 1 reports medians and
   ranges, not distributional shapes. The virtual cohort draws weight,
   albumin, age and baseline CRT from truncated normals centred on the
   published sex-specific medians with the range spanning about +/- 2 SD. The
   marginal medians and ranges therefore match Table 1, but the *correlation*
   structure between covariates does not: Figure S1 shows a correlated
   covariate matrix stratified by sex which is not reproducible from the
   published summaries. Any cohort-level percentile in this vignette should be
   read with that in mind; every published-value gate above is evaluated at
   fixed reference covariates and is unaffected.

2. **Spline slope denominators use nominal times.** Data S4 divides each
   spline segment by the difference of the *actual* observation times at the
   bracketing knots (`CKT1`..`CKT4`). Those equal the nominal 1, 2, 4 and 12 h
   whenever sampling is on schedule, and the model file writes them as the
   nominal values, since the actual per-subject times are not published.

3. **Baseline VAS beyond the last knot.** Data S4 assigns the flat value
   `100 * H4` only at exactly `NCKT == 12`, because no observation falls later.
   The model file holds the baseline flat at `100 * H4` for all times at or
   beyond 12 h, which is well-defined and agrees with the control stream
   everywhere the control stream is defined.

4. **Stirling versus exact gamma normalisation.** Data S1-S4 normalise the
   transit density with Stirling's approximation,
   `LNFAC = LOG(2.5066) + (NN + 0.5) * LOG(NN) - NN`. rxode2's builtin
   `transit()` evaluates the identical density but normalises with the exact
   `lgamma(NN + 1)`. At NN = 19.4 the two differ by roughly 0.4% of the input
   rate, and the exact form is the more correct of the two; the AUC gate above
   confirms that the whole dose is still delivered.

5. **Albumin unit conversion.** The canonical `ALB` register unit is SI g/L
   while Kim 2026 calibrated its coefficients against US-convention g/dL. Each
   model file applies the register-mandated inline conversion
   `alb_gdL <- ALB * 0.1` in `model()`, so the published coefficients are used
   unchanged and a model-data column of 44 g/L is the reference subject.

6. **Half-life is not tuned to the published NCA.** The model's disposition
   half-life is `r sprintf("%.2f", log(2) * 64 / 18)` h by construction, about
   18% shorter than the 2.9-3.1 h NCA mean quoted in the Introduction from the
   source clinical study. This follows from the published one-compartment
   structure and is left as-is.

## Convention deviations

None. All four models pass `checkModelConventions()` with only the
informational note that `units$dosing` (mg) and the `units$concentration`
numerator (ug) differ in magnitude -- the required 1000x scaling is applied
explicitly in each `model()` block as `Cc <- 1000 * central / vc`, matching the
control streams' `S2 = V2/1000`.

Three covariate canonicals were registered with this extraction: `CRT_BL`,
`DSST_BL` and `VAS_SEDATION_BL`.

```{r conventions, echo = FALSE}
for (m in c("Kim_2026_zolpidem", "Kim_2026_zolpidem_dsst",
            "Kim_2026_zolpidem_crt", "Kim_2026_zolpidem_vas")) {
  res <- nlmixr2lib::checkModelConventions(m)
  cat(m, ": ", sum(res$severity == "error"), " error(s), ",
      sum(res$severity == "warning"), " warning(s)\n", sep = "")
}
```
