---
title: "Infliximab TDMx model panel (Kim 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Infliximab TDMx model panel (Kim 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(PKNCA)
```

# Model and source

* Citation: Kim Y, Baek SH, Jang IJ, Chung JY. Model-Informed Precision Dosing
  of Infliximab in Korean Inflammatory Bowel Disease Patients: External
  Validation of Population Pharmacokinetic Models. CPT Pharmacometrics Syst
  Pharmacol. 2025;14:1682-1694.
* Article: [doi:10.1002/psp4.70089](https://doi.org/10.1002/psp4.70089)

Kim 2025 is an **external-validation** paper, not a model-development paper. It
evaluates the infliximab population PK models implemented in
[TDMx](https://www.tdmx.eu/), a free online model-informed precision dosing
(MIPD) dashboard, against a retrospective cohort of 199 Korean inflammatory
bowel disease (IBD) patients treated at Seoul National University Hospital
between 2019 and 2023 (142 adults with 294 concentrations, 57 children with 158
concentrations).

The paper itself develops no new model. What makes it extractable is its
supplement: `Data S1` (Supplementary Material 1) contains the **complete NONMEM
control stream for every model evaluated**, each run with `MAXEVAL=0` so that
all `$THETA`, `$OMEGA` and `$SIGMA` entries are held at the originally published
values. Kim 2025 Table S2 independently tabulates the same typical values and
inter-individual variability, giving a second, agreeing transcription of each
model. Every parameter in the model files extracted here is traced to that
control stream and cross-checked against Table S2.

Because the models are transcriptions of other authors' work, each extracted
file is named for the transcribing paper (`Kim_2025_infliximab_<model>`) and
credits the original developers in its `reference` field. This follows the
precedent already in the library: `modellib("Frymoyer_2017_infliximab")` is
likewise a second-hand transcription, taken from Frymoyer 2017's reproduction of
the Fasanmade model. **When the primary publications are extracted directly,
these files should be re-checked against them.**

## Which models are in this release

Kim 2025 evaluated three adult and seven paediatric models. Eight of the ten
control streams are extracted here -- every adult model, and five of the seven
paediatric ones. Two are deferred pending naming decisions that require new
canonical entries (see *Assumptions and deviations*).

| Kim 2025 stream | Original model | Extracted as | Status |
|---|---|---|---|
| #1 Aubourg (adult) | Aubourg 2015 | `Kim_2025_infliximab_aubourg` | included |
| #2 Passot (adult) | Passot 2016 | `Kim_2025_infliximab_passot_adult` | included |
| #3 Ternant (adult) | Ternant 2008 | `Kim_2025_infliximab_ternant` | included |
| #4 Bauman (paediatric) | Bauman 2020 | -- | deferred |
| #5 Dubinsky (paediatric) | Dubinsky 2017 | `Kim_2025_infliximab_dubinsky` | included |
| #6 Fasanmade combined | Fasanmade 2011 | `Kim_2025_infliximab_fasanmade_combined` | included |
| #7 Fasanmade paediatric | Fasanmade 2011 | `Kim_2025_infliximab_fasanmade_pediatric` | included |
| #8 Passot (paediatric) | Passot 2016 | `Kim_2025_infliximab_passot_pediatric` | included |
| #9 Petitcollin (paediatric) | Petitcollin 2018 | -- | deferred |
| #10 Wojciechowski | Wojciechowski 2017 | `Kim_2025_infliximab_wojciechowski` | included |

Kim 2025 additionally lists the Dreesen and Xiong models as implemented in TDMx
but excluded them from its own analysis, because the validation dataset lacked
faecal calprotectin, Crohn's Disease Activity Index and neutrophil CD64 activity
ratio. No control stream is provided for either, so neither is extractable here.

## Headline findings of the paper

* All ten models showed a **consistent positive bias** (over-prediction): the
  95% CI of the mean percentage error excluded zero for every model, and every
  MPE exceeded the +/-20% clinical-acceptability threshold.
* Among adult models the **Passot model performed best** (MPE 26.4%, MAE
  1.1 mg/L, rRMSE 159.0%) and was the only adult model whose normalised
  prediction distribution errors were acceptable. It is the model the authors
  recommend for MIPD in Korean adults.
* **No paediatric model was clinically acceptable**, even after supplying one or
  two previous trough concentrations (MPE 30.4%-143.4%, MAE 1.4-2.6 mg/L, rRMSE
  96.3%-564.0%).
* Assuming every patient to be ATI-positive *improved* predictive performance,
  which the authors attribute to the models' inherent positive bias.

# Population

The development population of each extracted model, as summarised by Kim 2025
Table S1:

```{r population-table}
pop_tab <- tibble::tribble(
  ~Model,                 ~`IBD type (n)`,   ~Population,        ~`ATI+ (%)`, ~`Age (y)`, ~`Female (%)`, ~`Weight (kg)`, ~`ALB (g/dL)`,
  "Aubourg",              "CD (133)",        "Adults",           "0",         "NS",       "59",          "60",           "NS",
  "Passot (both)",        "CD (63), UC (16)", "Paediatric+adult", "0",        "NS",       "NS",          "NS",           "NS",
  "Ternant",              "CD (30), UC (3)", "Adults",           "15",        "33",       "55",          "67",           "NS",
  "Dubinsky",             "CD (41), UC (9)", "Paediatric",       "28",        "13",       "44",          "41",           "3.9",
  "Fasanmade combined",   "CD (692)",        "Paediatric+adult", "10",        "33",       "56",          "65",           "4.1",
  "Fasanmade paediatric", "CD (112)",        "Paediatric",       "3",         "13",       "41",          "42",           "3.8",
  "Wojciechowski",        "CD (112), UC (543)", "Paediatric+adult", "NS",     "NS",       "NS",          "70",           "4.0"
)
knitr::kable(pop_tab, caption = "Development populations (Kim 2025 Table S1). NS = not specified. Medians, except Dubinsky (means).")
```

The Kim 2025 **validation** cohort (Table 1) was 142 adults (age 31.9 +/- 11.4 y,
weight 66.7 +/- 14.9 kg, 24.6% female, albumin 4.4 +/- 0.5 g/dL) and 57 children
(age 14.2 +/- 2.8 y, weight 58.7 +/- 16.3 kg, 26.3% female, albumin
4.1 +/- 0.5 g/dL), all East Asian, 80.3% / 87.7% Crohn's disease. Observed trough
concentrations had a median of 3.3 mg/L in both groups (range 0.1-24.8 mg/L in
adults, 0.1-46.8 mg/L in children). That cohort was used only to assess
predictive performance -- it did not update any parameter.

# Source trace

Every value in every extracted model file comes from the `Data S1` control
stream, cross-checked against Table S2.

| Item | Source location |
|---|---|
| Structural `$THETA` (CL, Vc, Vp, Q) | Data S1 Supplementary Material 1, per-model `$THETA` block; Table S2 "Parameter values" columns |
| Covariate exponents / coefficients | Same `$THETA` block, per-model comments naming each coefficient |
| Covariate functional forms | Per-model `$PK` block |
| ODE structure | Per-model `$MODEL` and `$DES` blocks |
| Residual error form | Per-model `$ERROR` block (`Y = IPRED*(1+ERR(1))` or `+ ERR(2)`) |
| Residual error magnitudes | Per-model `$SIGMA` block; the SDs are given in the stream's own comments |
| IIV variances | Per-model `$OMEGA` block; %CV cross-checked against Table S2 |
| IIV scale convention | Aubourg stream annotation: `$OMEGA ; Interindividual standard deviation OMEGA = (%CV/100)**2` |
| Development populations | Table S1 |
| Model performance (MPE / MAE / rRMSE) | Table 2 |

The IIV scale deserves a note, because it is the usual ambiguity in a
transcription of this kind. Kim 2025's streams use `OMEGA = CV^2` rather than the
exact log-normal `omega^2 = log(1 + CV^2)`. This is verifiable independently in
every model: for example Dubinsky's `$OMEGA` values 0.098 / 0.010 / 0.579 / 1.232
are exactly `0.313^2` / `0.0985^2` / `0.761^2` / `1.11^2`, and Table S2 reports
those same figures as 31.3% / 9.85% / 76.1% / 111%. The variances are carried
through unchanged.

# Load the models

```{r load-models}
model_names <- c(
  "Kim_2025_infliximab_aubourg",
  "Kim_2025_infliximab_passot_adult",
  "Kim_2025_infliximab_ternant",
  "Kim_2025_infliximab_passot_pediatric",
  "Kim_2025_infliximab_dubinsky",
  "Kim_2025_infliximab_wojciechowski",
  "Kim_2025_infliximab_fasanmade_combined",
  "Kim_2025_infliximab_fasanmade_pediatric"
)

# readModelDb() returns the model FUNCTION; rxode2::rxode() resolves it to a ui.
models <- lapply(model_names, function(nm) rxode2::rxode(readModelDb(nm)))
names(models) <- model_names

tibble::tibble(
  Model = sub("^Kim_2025_infliximab_", "", model_names),
  Compartments = vapply(models, function(ui) length(ui$state), integer(1)),
  Covariates = vapply(models, function(ui) paste(sort(ui$allCovs), collapse = ", "), character(1))
) |>
  knitr::kable(caption = "Structure of the extracted models.")
```

# Virtual cohort and simulation helpers

Two reference subjects are used throughout: a 65 kg adult and a 40 kg child,
both male, Crohn's disease, ADA-negative, no concomitant immunomodulator,
serum albumin 40 g/L (= 4.0 g/dL). These match the central tendency of the
Kim 2025 validation cohort (Table 1) closely enough to make the models
comparable to one another.

```{r cohort-def}
ref_cov <- list(
  adult = list(WT = 65, SEXF = 0, IBD_CD = 1, AGE = 32, ALB = 40,
               ADA_POS = 0, CONMED_IMMUNOMOD = 0),
  child = list(WT = 40, SEXF = 0, IBD_CD = 1, AGE = 13, ALB = 40,
               ADA_POS = 0, CONMED_IMMUNOMOD = 0)
)

# Dense early sampling so that the distribution phase is resolved; PKNCA needs a
# time-zero record, which seq(0, ...) supplies.
obs_times <- unique(c(
  seq(0, 0.5, by = 1 / 24),
  seq(0.75, 7, by = 0.25),
  seq(7.5, 56, by = 0.5)
))

# Attach only the covariates a given model actually declares, so rxode2 does not
# warn about unused columns.
attach_covariates <- function(dat, ui, cov) {
  for (nm in intersect(names(cov), ui$allCovs)) {
    dat[[nm]] <- cov[[nm]]
  }
  dat
}

# Single 5 mg/kg IV infusion over 2 h (0.0833 day), observed on the ODE state.
single_dose_events <- function(ui, cov) {
  ev <- rxode2::et(amt = 5 * cov$WT, dur = 0.0833, cmt = "central")
  ev <- rxode2::et(ev, obs_times, cmt = "central")
  attach_covariates(as.data.frame(ev), ui, cov)
}

solve_typical <- function(ui, dat) {
  out <- rxode2::rxSolve(rxode2::zeroRe(ui), dat, returnType = "data.frame")
  if (is.null(out$id)) out$id <- 1L
  out
}
```

# Typical-value profiles across the model panel

```{r typical-profiles, fig.width = 7, fig.height = 4.5}
typical <- bind_rows(lapply(model_names, function(nm) {
  bind_rows(lapply(names(ref_cov), function(grp) {
    ui <- models[[nm]]
    dat <- single_dose_events(ui, ref_cov[[grp]])
    solve_typical(ui, dat) |>
      filter(!is.na(Cc)) |>
      transmute(model = sub("^Kim_2025_infliximab_", "", nm),
                cohort = grp, time, Cc, cl, vc)
  }))
}))

ggplot(typical, aes(time, Cc, colour = model)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~cohort, labeller = labeller(
    cohort = c(adult = "Adult, 65 kg, 325 mg", child = "Child, 40 kg, 200 mg"))) +
  scale_y_log10() +
  labs(x = "Time (days)", y = "Infliximab concentration (mg/L)",
       colour = "Model",
       title = "Typical-value profiles after a single 5 mg/kg IV infusion") +
  theme_bw() +
  theme(legend.position = "bottom")
```

The panel spread is itself the paper's point: these models were all fitted to
infliximab in IBD, yet at an identical dose in an identical subject their typical
profiles differ severalfold in the terminal phase. That is the variability an
MIPD dashboard inherits when it offers a choice of prior.

# Structural validation: AUC(0-inf) = Dose / CL

For any linear model, the exact identity `AUC(0-inf) = Dose / CL` holds. This is
the strictest available check on each transcription, because it ties the
simulated exposure to the published clearance and to the covariate model that
scales it. It is asserted per model and per cohort.

```{r pknca-setup}
conc_dat <- typical |>
  transmute(model, cohort,
            treatment = paste(model, cohort, sep = " / "),
            id = 1L, time, Cc, cl)

dose_dat <- conc_dat |>
  group_by(model, cohort, treatment, id) |>
  summarise(time = 0, .groups = "drop") |>
  left_join(
    tibble::tibble(cohort = c("adult", "child"),
                   dose = c(5 * ref_cov$adult$WT, 5 * ref_cov$child$WT)),
    by = "cohort"
  )

# PKNCAdose() rejects a nested (slash) grouping formula; use the same
# `| treatment + id` grouping on BOTH objects.
o_conc <- PKNCA::PKNCAconc(conc_dat |> filter(!is.na(Cc)), Cc ~ time | treatment + id)
o_dose <- PKNCA::PKNCAdose(dose_dat, dose ~ time | treatment + id)

o_data <- PKNCA::PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(
    start = 0, end = 56,
    cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE
  )
)
res <- PKNCA::pk.nca(o_data)
nca <- as.data.frame(res)
```

```{r identity-check}
expected <- conc_dat |>
  group_by(model, cohort, treatment) |>
  summarise(cl = first(cl), .groups = "drop") |>
  left_join(tibble::tibble(cohort = c("adult", "child"),
                           dose = c(5 * ref_cov$adult$WT, 5 * ref_cov$child$WT)),
            by = "cohort") |>
  mutate(auc_expected = dose / cl)

identity_tab <- nca |>
  filter(PPTESTCD == "aucinf.obs") |>
  select(treatment, aucinf_sim = PPORRES) |>
  left_join(expected, by = "treatment") |>
  mutate(ratio = aucinf_sim / auc_expected,
         pct_diff = 100 * (ratio - 1))

identity_tab |>
  transmute(Model = model, Cohort = cohort,
            `CL (L/day)` = round(cl, 4),
            `Dose/CL (mg/L*day)` = round(auc_expected, 1),
            `Simulated AUCinf` = round(aucinf_sim, 1),
            `Difference (%)` = round(pct_diff, 2)) |>
  knitr::kable(caption = "AUC(0-inf) identity check against Dose/CL.")

# Strict assertion: NCA-derived AUCinf must reproduce Dose/CL to within 1%.
stopifnot(nrow(identity_tab) == length(model_names) * length(ref_cov))
stopifnot(all(abs(identity_tab$pct_diff) < 1))
```

Every model reproduces `Dose / CL` to well within 1%, confirming that the
clearance value, the covariate model scaling it, the compartment structure and
the dose-to-concentration unit chain are all internally consistent.

# Comparison against the published parameter values

The paper reports no NCA summary (Cmax / AUC / half-life) for any model, so the
published quantities available for comparison are the typical parameter values in
Table S2. The check below recomputes each model's clearance at its **own
reference covariates** -- where every covariate multiplier collapses to one, or to
the value the source stream dictates -- and compares the simulated exposure with
the exposure implied by the Table S2 clearance.

```{r published-comparison}
# Table S2 typical CL, and the covariate values at which that CL applies.
published_cl <- tibble::tribble(
  ~model,                 ~cl_published, ~WT, ~ALB_gdL,
  "aubourg",              0.456,         60,  NA,     # male typical CL; Vc reference weight 60 kg
  "passot_adult",         0.23,          67,  NA,     # before the disease/sex multipliers
  "ternant",              0.288,         67,  NA,     # ADA-negative typical CL; Vc reference weight 67 kg
  "passot_pediatric",     0.23,          67,  NA,
  "dubinsky",             0.296,         70,  4.0,
  "wojciechowski",        0.294,         70,  4.0,
  "fasanmade_combined",   0.3523,        65,  4.1,    # 5.42 mL/kg/day x 65 kg / 1000
  "fasanmade_pediatric",  0.22806,       42,  3.8     # 5.43 mL/kg/day x 42 kg / 1000
)

ref_cl <- bind_rows(lapply(seq_len(nrow(published_cl)), function(i) {
  row <- published_cl[i, ]
  nm <- paste0("Kim_2025_infliximab_", row$model)
  ui <- models[[nm]]
  cov <- list(
    WT = row$WT, SEXF = 0, IBD_CD = 1, AGE = 32,
    ALB = if (is.na(row$ALB_gdL)) 40 else row$ALB_gdL * 10,
    ADA_POS = 0, CONMED_IMMUNOMOD = 0
  )
  dat <- single_dose_events(ui, cov)
  out <- solve_typical(ui, dat)
  tibble::tibble(model = row$model,
                 cl_model = out$cl[1],
                 cl_published = row$cl_published)
}))

ref_cl |>
  mutate(`Ratio` = round(cl_model / cl_published, 4),
         `Difference (%)` = round(100 * (cl_model / cl_published - 1), 2)) |>
  transmute(Model = model,
            `Published CL (L/day)` = cl_published,
            `Model CL at reference (L/day)` = round(cl_model, 5),
            Ratio, `Difference (%)`) |>
  knitr::kable(caption = "Clearance at each model's own reference covariates vs Kim 2025 Table S2.")
```

Aubourg, Ternant, Dubinsky, Wojciechowski, and both Fasanmade parameterizations
reproduce the published clearance exactly at their reference covariates.
(Ternant's row is its ADA-negative stratum; the ADA-positive stratum is a
separate typical value of 0.768 L/day, exercised by the panel simulations
above only if `ADA_POS` is set to 1.) The two Passot
rows differ by construction and are the informative case: Passot's model was
fitted across several chronic inflammatory diseases, so **neither IBD subtype is
a unit multiplier**. A Crohn's disease patient carries `exp(0.384) = 1.468` and an
ulcerative colitis patient `exp(0.472) = 1.603` on top of the tabulated 0.23
L/day. The 0.23 in Table S2 is therefore a base value that no actual IBD patient
ever has, which is exactly the kind of detail a summary table loses and a control
stream preserves.

```{r nca-comparison-table}
# The paper publishes no NCA parameters, so the reference AUC below is derived
# from the Table S2 clearance (AUCinf = Dose/CL), evaluated at the reference
# covariates used in the identity check above.
simulated_nca <- nca |>
  filter(PPTESTCD %in% c("cmax", "aucinf.obs", "half.life")) |>
  left_join(conc_dat |> distinct(treatment, model, cohort), by = "treatment") |>
  filter(cohort == "adult") |>
  select(model, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

reference_nca <- expected |>
  filter(cohort == "adult") |>
  transmute(model, aucinf.obs = auc_expected)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = simulated_nca,
  reference = reference_nca,
  by = "model",
  params = "aucinf.obs",
  tolerance_pct = 20,
  label_first_column = "NCA parameter"
)
knitr::kable(cmp, caption = "Simulated AUC(0-inf) vs the exposure implied by the published clearance (adult, 325 mg).")
```

# Inter-individual variability: the paper's paediatric-model claim

Kim 2025 reports that the paediatric models carry much larger IIV than the adult
models, with Dubinsky and Wojciechowski reaching roughly 110%, and argues that
such large IIV "may obscure models' structural deficiencies, giving a misleading
impression of good model fit despite poor data characterization". The extracted
`$OMEGA` values should reproduce that ordering.

```{r iiv-check}
iiv_tab <- bind_rows(lapply(model_names, function(nm) {
  ui <- models[[nm]]
  d <- ui$iniDf
  d <- d[!is.na(d$neta1) & d$neta1 == d$neta2, ]
  tibble::tibble(model = sub("^Kim_2025_infliximab_", "", nm),
                 eta = d$name, variance = d$est, cv_pct = 100 * sqrt(d$est))
}))

iiv_tab |>
  mutate(cv_pct = round(cv_pct, 1)) |>
  pivot_wider(id_cols = model, names_from = eta, values_from = cv_pct) |>
  knitr::kable(caption = "IIV as %CV (sqrt of the published OMEGA variance).")

# The paper's claim: Dubinsky and Wojciechowski carry ~110% IIV on Q, far above
# anything in the adult models. Ternant also carries an etalq (at 10% CV), so
# the assertion names the models rather than resting on the eta name alone.
q_iiv <- iiv_tab |> filter(eta == "etalq")
stopifnot(setequal(q_iiv$model, c("dubinsky", "wojciechowski", "ternant")))

big_q <- q_iiv |> filter(cv_pct > 105)
stopifnot(setequal(big_q$model, c("dubinsky", "wojciechowski")))
stopifnot(all(big_q$cv_pct < 115))

# No eta of any of the three adult models exceeds 50% CV.
max_adult_iiv <- iiv_tab |>
  filter(model %in% c("aubourg", "passot_adult", "ternant")) |>
  summarise(m = max(cv_pct)) |>
  pull(m)
stopifnot(max_adult_iiv < 50)
```

The assertion holds: the only two etas above 105% CV in the panel are the
inter-compartmental clearance terms of Dubinsky (111%) and Wojciechowski (110%),
while no adult-model eta exceeds 50%.

# Steady-state troughs against the observed distribution

The clinically relevant quantity in this paper is the maintenance trough. Kim
2025 reports observed troughs with a median of 3.3 mg/L in both cohorts, against
a widely used therapeutic target of >5 mg/L, and finds that all the models
over-predict. Simulating the standard 5 mg/kg every-8-weeks maintenance regimen
shows where each model's typical trough sits relative to that observed median.

```{r trough-sim, fig.width = 7, fig.height = 4}
maint_trough <- bind_rows(lapply(model_names, function(nm) {
  bind_rows(lapply(names(ref_cov), function(grp) {
    ui <- models[[nm]]
    cov <- ref_cov[[grp]]
    ev <- rxode2::et(amt = 5 * cov$WT, dur = 0.0833, ii = 56, addl = 8,
                     cmt = "central")
    ev <- rxode2::et(ev, seq(0, 56 * 9, by = 1), cmt = "central")
    dat <- attach_covariates(as.data.frame(ev), ui, cov)
    out <- solve_typical(ui, dat) |> filter(!is.na(Cc))
    tibble::tibble(model = sub("^Kim_2025_infliximab_", "", nm),
                   cohort = grp,
                   trough = out$Cc[out$time == 56 * 9])
  }))
}))

ggplot(maint_trough, aes(reorder(model, trough), trough, fill = cohort)) +
  geom_col(position = position_dodge()) +
  geom_hline(yintercept = 3.3, linetype = "dashed") +
  geom_hline(yintercept = 5, linetype = "dotted") +
  coord_flip() +
  labs(x = NULL, y = "Steady-state trough (mg/L)", fill = "Cohort",
       title = "Typical steady-state trough, 5 mg/kg q8w",
       subtitle = "Dashed = observed median 3.3 mg/L (Kim 2025 Table 1); dotted = 5 mg/L target") +
  theme_bw()

knitr::kable(
  maint_trough |>
    mutate(trough = round(trough, 2)) |>
    pivot_wider(names_from = cohort, values_from = trough),
  caption = "Typical steady-state trough (mg/L) after 5 mg/kg q8w."
)
```

Every model's typical trough falls below the 5 mg/L target at 5 mg/kg q8w, which
is consistent with the clinical picture the paper describes: Korean patients were
sampled reactively, on suspicion of loss of response, and the observed median
trough was 3.3 mg/L. Note that the paper's "positive bias" finding is a statement
about *individual* predictions against *that specific reactive-sampling cohort*,
not something reproducible from typical-value simulation alone -- the validation
metrics in Table 2 depend on patient-level data that is not distributed with the
paper.

# A discrepancy with `Frymoyer_2017_infliximab`

The library already contains an independent transcription of the same underlying
Fasanmade REACH + ACCENT I model, taken from Frymoyer 2017. The two agree on all
four typical values (5.42 mL/kg/day, 52.4 mL/kg, 19.6 mL/kg, 2.26 mL/kg/day), on
the albumin exponent (-0.855), and on the ADA (+29%) and immunomodulator (-13.7%)
effects. They disagree on the **sign of the three body-weight exponents**.

```{r frymoyer-discrepancy, fig.width = 7, fig.height = 4}
wt_grid <- seq(20, 110, by = 1)
kim_cl <- 0.3523 * (wt_grid / 65)^(1 - 0.313)       # Kim 2025 Data S1 model #6
fry_cl <- 0.3523 * (wt_grid / 65)^(1 + 0.313)       # Frymoyer_2017_infliximab

tibble::tibble(
  WT = rep(wt_grid, 2),
  CL = c(kim_cl, fry_cl),
  source = rep(c("Kim 2025 (exponent 0.687)", "Frymoyer 2017 (exponent 1.313)"),
               each = length(wt_grid))
) |>
  ggplot(aes(WT, CL, colour = source)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = 65, linetype = "dashed") +
  labs(x = "Body weight (kg)", y = "Typical CL (L/day)", colour = NULL,
       title = "Fasanmade REACH + ACCENT I: two transcriptions of the weight effect",
       subtitle = "Identical at the 65 kg reference (dashed), divergent away from it") +
  theme_bw() +
  theme(legend.position = "bottom")

tibble::tibble(
  `Body weight (kg)` = c(30, 45, 65, 90),
  `Kim 2025 CL` = round(0.3523 * (c(30, 45, 65, 90) / 65)^(1 - 0.313), 4),
  `Frymoyer 2017 CL` = round(0.3523 * (c(30, 45, 65, 90) / 65)^(1 + 0.313), 4)
) |>
  mutate(`Difference (%)` = round(100 * (`Frymoyer 2017 CL` / `Kim 2025 CL` - 1), 1)) |>
  knitr::kable(caption = "Divergence between the two transcriptions.")
```

The two encodings coincide exactly at 65 kg and diverge away from it -- by about
-38% at 30 kg and +19% at 90 kg. Kim 2025's control stream is explicit
(`CLWGT = (WGT/65)**THETA(8)` with `THETA(8) = -0.313`, applied on top of
`TVCL = THETA(1)*WGT/1000`), which yields a **total**-clearance weight exponent of
0.687 -- close to the canonical allometric 0.75. The Frymoyer transcription
implies a total exponent of 1.313, i.e. clearance rising faster than
proportionally with body weight, which is not a plausible allometric relationship
for a monoclonal antibody. These files follow Kim 2025.

This is flagged rather than resolved here: settling it requires the Fasanmade
2011 primary publication, which is not on disk. `Frymoyer_2017_infliximab` is left
untouched by this extraction.

# Assumptions and deviations

* **Second-hand transcription.** Every model here is Kim 2025's re-coding of
  another author's model, not a direct extraction from the primary publication.
  Kim 2025 states it confirmed agreement between its NONMEM implementations and
  TDMx (Figure S3), which is a meaningful but not independent check. When the
  primary papers are extracted, these files should be re-verified against them.
* **Two models deferred.** Bauman and Petitcollin are not included, because each
  needs a canonical name that does not yet exist in the library registers and
  that requires an explicit naming decision:
  * *Bauman* uses erythrocyte sedimentation rate as a covariate on clearance
    (`(ESR/9)^0.109`). `ESR` has no entry in `inst/references/covariate-columns.md`,
    and no model in the library currently uses it as a covariate column.
  * *Petitcollin* models immunisation as a logistic-in-time multiplier on
    clearance (`clvar^(expit(-baserisk + beta*t))`). This is a third form of
    time-varying clearance beyond the `cl_hill_` and `cl_exp_` stems registered
    in `references/parameter-names.md`, and `checkModelConventions()` warns on a
    clearance expression that references time without one of them.

  Both are otherwise fully specified by their control streams, so each becomes a
  one-file addition once its name is ratified. Neither omission is a
  data-availability gap.
* **Ternant's non-standard structure.** Ternant is the one model in the panel
  that departs from the panel's usual multiplicative covariate model in two
  ways, both reproduced verbatim:
  * *Clearance is stratified, not offset.* A wholly separate typical CL **and a
    separate CL eta** are estimated in each ADA stratum (0.288 vs 0.768 L/day),
    so the file uses the stratum-suffix convention (`lcl_ada_neg` /
    `lcl_ada_pos`) rather than a reference value plus `e_ada_cl`.
  * *The central volume is an additive decomposition.* `Vc = 1.7*(WT/67) +
    [1.1 female | 2.3 male]` -- a weight-proportional term plus a
    weight-independent sex intercept, each with its own eta, and linear in
    `WT/67` rather than a power function. The three components are named
    `lvc_wt`, `lvc_female` and `lvc_male` by analogy with the registered
    additive multi-component clearance stems (`lcl_renal` / `lcl_nonren`); the
    additive-volume case is not itself registered.

  The result is seven random effects fitted to 33 patients, the smallest
  development cohort in the panel -- worth bearing in mind when using it as a
  Bayesian prior. Kim 2025 found it gave the best MAE (0.8 mg/L) but the worst
  rRMSE (294.5%) of the three adult models.
* **IIV scale.** Kim 2025's streams use `OMEGA = CV^2`, not the exact log-normal
  `omega^2 = log(1 + CV^2)`. The published variances are carried through
  unchanged rather than being converted, because the control stream is the
  authority for what was actually run.
* **Inter-occasion variability.** Both Fasanmade streams declare two IOV
  occasions on clearance but hard-code `FLAG1 = 1, FLAG2 = 0` ("In this cohort,
  all data were from the maintenance phase"), so exactly one occasion is ever
  active. Only that occasion's random effect is carried (`etaiov_cl_1`, following
  the registry's established IOV naming); the second, structurally inert
  occasion is not declared. It is routed through a named intermediate
  (`iov_cl`) in `model()` so that the mu-referenced exponent holds a single eta,
  which rxode2 requires.
* **Albumin units.** The canonical `ALB` column is SI g/L; every source stream is
  calibrated on g/dL. Each model file converts inline (`alb_gdL <- ALB * 0.1`).
* **`IBD_CD` polarity.** The source streams code `DX = 0` for Crohn's disease and
  `DX = 1` for ulcerative colitis, the inverse of the canonical `IBD_CD`. Derive
  `IBD_CD = 1 - DX`.
* **Stray accumulator.** The Passot paediatric stream declares one compartment but
  carries a `DADT(2) = A(1)/V1` line. Compartment 2 is never declared in `$MODEL`
  and never enters `$ERROR`, so it is an unused cumulative-AUC accumulator; it is
  not reproduced.
* **No published NCA to compare against.** The paper reports no Cmax / AUC /
  half-life for any model, so the NCA comparison above is against the exposure
  implied by the published clearance (`Dose/CL`) rather than against a published
  NCA table. The model-performance metrics in Table 2 (MPE / MAE / rRMSE) depend
  on the retrospective patient-level dataset, which is not distributed.
* **Reference covariates.** The 65 kg adult and 40 kg child used here are chosen
  to sit near the Kim 2025 validation cohort's central tendency (Table 1); they
  are not prescribed by the paper.
