---
title: "Nifedipine + captopril cardiovascular PK-PD in spontaneously hypertensive rats (Kiriyama 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nifedipine + captopril cardiovascular PK-PD in spontaneously hypertensive rats (Kiriyama 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

This paper contributes **two** model files, one per drug, because Kiriyama and
colleagues fitted nifedipine and captopril independently (each drug's PK from its
own monotherapy arm, each drug's PD from its own monotherapy arm) and then
*predicted* the coadministration response by superposing the two separately
fitted effect models. Reproducing that structure keeps the paper's central
claim -- that the observed combination response differs from the additive
prediction -- testable.

* Nifedipine: `r rxode2::rxode2(readModelDb("Kiriyama_2024_nifedipine_rat"))$description`
* Captopril: `r rxode2::rxode2(readModelDb("Kiriyama_2024_captopril_rat"))$description`
* Citation: `r rxode2::rxode2(readModelDb("Kiriyama_2024_nifedipine_rat"))$reference`
* Article: <https://doi.org/10.1002/prp2.1249>

```{r load-models}
nif <- rxode2::rxode2(readModelDb("Kiriyama_2024_nifedipine_rat"))
cap <- rxode2::rxode2(readModelDb("Kiriyama_2024_captopril_rat"))
```

# Population

Male spontaneously hypertensive rats (SHR/Izm, 215-315 g) were anaesthetised with
intraperitoneal urethane (1.0 g/kg), catheterised in the left carotid artery for
blood pressure and instrumented for ECG. Blood pressure (BP), heart rate (HR) and
QT interval were recorded continuously from -15 to 180 min after the start of a
30-minute intravenous infusion into the right femoral vein; plasma was sampled
from the right jugular vein between 5 and 300 min and assayed by LC-MS/MS
(Methods sections 2.2-2.3).

Monotherapy doses were nifedipine 1.0 mg/kg and captopril 15.0 mg/kg. For
coadministration both doses were reduced -- nifedipine 0.5 mg/kg and captopril
5.0 mg/kg -- chosen so that the combination would produce an effect comparable to
either drug alone. Group sizes differ by analysis: the PK parameters of Tables 1
and 2 come from 3-6 rats, the PD parameters of Table 3 from 8-10 rats. The paper
reports that baseline BP, HR and QT did not differ significantly across the
nifedipine, captopril and coadministration groups.

The same information is available programmatically via each model's `population`
metadata (`readModelDb("Kiriyama_2024_nifedipine_rat")()$population`).

## Dose scale: recovering the body weight

The paper reports doses in mg/kg but fits **absolute** central volumes (mL) and
absolute micro-constants, so converting a dose to an event-table `amt` requires a
body weight that the paper never states beyond the 215-315 g range. It is
recoverable: the Table 2 parameters plus the Figure 2 fitted peak concentrations
pin it down, and the two drugs give independent estimates.

```{r weight-backsolve}
# Peak of the fitted (solid) monotherapy curves read from Figure 2.
fig2_peak <- c(nifedipine = 2900, captopril = 19500)

peak_per_mg <- function(mod, pars) {
  ev <- rbind(
    data.frame(time = 0, amt = 1e6, dur = 30, evid = 1L,
               cmt = "central", dvid = NA_integer_),
    data.frame(time = seq(0, 60, by = 1), amt = NA_real_, dur = NA_real_,
               evid = 0L, cmt = "central", dvid = 1L)
  )
  s <- rxode2::rxSolve(mod, ev, params = pars, returnType = "data.frame",
                       useLinCmt = FALSE, atol = 1e-12, rtol = 1e-12)
  s$Cc[s$time == 30]   # ng/mL per 1 mg infused over 30 min
}

nif_pk_mono <- c(lvc = log(54.9), lkel = log(0.0218), lk12 = log(0.0121), lk21 = log(0.0165))
cap_pk_mono <- c(lvc = log(54.2), lkel = log(0.0521), lk12 = log(0.1530), lk21 = log(0.0694))

implied_g <- c(
  nifedipine = 1000 * (fig2_peak[["nifedipine"]] / peak_per_mg(nif, nif_pk_mono)) / 1.0,
  captopril  = 1000 * (fig2_peak[["captopril"]]  / peak_per_mg(cap, cap_pk_mono)) / 15.0
)
round(implied_g, 1)
```

Both drugs independently imply a rat of about 250 g -- agreeing to within 1% of
each other and comfortably inside the reported 215-315 g range. Every simulation
below therefore doses a 250 g rat.

```{r weight-gate}
# Two independent estimates must agree; if they diverge the dose scale, the PK
# parameters, or the ODE encoding is wrong.
stopifnot(
  abs(diff(implied_g)) / mean(implied_g) < 0.02,
  all(implied_g > 215, implied_g < 315)
)
WT <- 0.250  # kg
```

# Source trace

Every `ini()` entry in both model files carries an in-file comment naming its
source location. They are collected here for review.

| Model | Equation / parameter | Value | Source location |
|---|---|---|---|
| both | `d/dt(central)`, `d/dt(peripheral1)` (2-compartment, zero-order infusion) | n/a | Methods 2.4 equations; Figure 1 |
| both | `d/dt(effect_*)` = `ke0 * (Cc - effect)` | n/a | Methods 2.5, `dC3/dt` equation |
| both | `hr`, `qt`, `qtc` sigmoid Emax | n/a | Methods 2.5, `E = E0 -/+ Emax*C3^gamma/(EC50^gamma + C3^gamma)`; Figure 1B |
| nifedipine | `bp` (hypotensive + EHS hypertensive arms) | n/a | Methods 2.5 `dE/dt` equation; Figure 1A |
| nifedipine | `d/dt(ehs)` | n/a | Methods 2.5 `dC4/dt` equation (see Errata) |
| nifedipine | `lvc` / `lkel` / `lk12` / `lk21` | 54.9 mL / 0.0218 / 0.0121 / 0.0165 min<sup>-1</sup> | Table 2, nifedipine Monotherapy |
| nifedipine | `lke0_bp`, `le0_bp`, `lemax_bp`, `lec50_bp` | 0.135 min<sup>-1</sup>, 96.8, 67.1 mmHg, 86.3 ng/mL | Table 3, nifedipine BP |
| nifedipine | `lkin_ehs`, `lkout_ehs`, `lgamma_bp` | 0.0217 unit/min, 0.0119 min<sup>-1</sup>, 1.17 | Table 3, nifedipine BP (K<sub>EHS,in</sub>, k<sub>EHS,out</sub>, alpha) |
| nifedipine | `lemax_bp_ehs`, `lec50_bp_ehs` | 49 338 mmHg, 1450 unit | Table 3, nifedipine BP (E<sub>max,2</sub>, EC<sub>50,2</sub>) |
| nifedipine | `lke0_hr`, `le0_hr`, `lemax_hr`, `lec50_hr`, `lhill_hr` | 0.006589 min<sup>-1</sup>, 306, 63.3 beat/min, 115 ng/mL, 0.931 | Table 3, nifedipine HR |
| nifedipine | `lke0_qt`, `le0_qt`, `lemax_qt`, `lec50_qt`, `lhill_qt` | 0.0117 min<sup>-1</sup>, 75.8, 76.3 msec, 3227 ng/mL, 0.857 | Table 3, nifedipine QT |
| nifedipine | `lke0_qtc`, `le0_qtc`, `lemax_qtc`, `lec50_qtc`, `lhill_qtc` | 0.0130 min<sup>-1</sup>, 169, 29.5 msec, 522 ng/mL, 1.92 | **Results prose** (Figure 3D paragraph), not Table 3 |
| captopril | `lvc` / `lkel` / `lk12` / `lk21` | 54.2 mL / 0.0521 / 0.153 / 0.0694 min<sup>-1</sup> | Table 2, captopril Monotherapy |
| captopril | `lke0_bp`, `le0_bp`, `lemax_bp`, `lec50_bp`, `lhill_bp` | 0.0400 min<sup>-1</sup>, 105, 41.8 mmHg, 7885 ng/mL, 0.662 | Table 3, captopril BP |
| captopril | `lke0_hr`, `le0_hr`, `lemax_hr`, `lec50_hr`, `lhill_hr` | 1.21e-5 min<sup>-1</sup>, 312.8, 80.4 beat/min, 7.6 ng/mL, 1.4 | Table 3, captopril HR |
| captopril | `lke0_qt`, `le0_qt`, `lemax_qt`, `lec50_qt`, `lhill_qt` | 4.05e-5 min<sup>-1</sup>, 81.2, 41.9 msec, 4873 ng/mL, 0.1347 | Table 3, captopril QT |
| both | coadministration PK override (`lvc`/`lkel`/`lk12`/`lk21`) | nifedipine 26.7 mL, 0.0768, 0.132, 0.0764; captopril 41.7 mL, 0.0842, 0.0631, 0.0529 | Table 2, Combination columns |
| both | residual error and IIV | not reported (`fixed(0)`, no etas) | Methods 2.4-2.5 (WinNonlin fits to mean profiles; only AIC reported) |

# Virtual cohort

The models are deterministic: the paper fitted mean profiles by naive pooling in
WinNonlin and reported no inter-individual variability and no residual error, so
neither model file carries `eta` terms. Each arm is therefore a single typical
250 g rat rather than a sampled cohort, and no `zeroRe()` / `omega = NA` step is
needed (there is nothing to zero). This is far below the 200-per-arm cap.

```{r cohort}
# Table 2 parameter sets. The paper reports a separate PK fit for each drug under
# monotherapy and under coadministration; it applies them by scenario rather than
# as a covariate effect, because it concludes that none of the PK differences
# reached statistical significance.
pk_sets <- list(
  nif_mono  = nif_pk_mono,
  nif_combo = c(lvc = log(26.7), lkel = log(0.0768), lk12 = log(0.132),  lk21 = log(0.0764)),
  cap_mono  = cap_pk_mono,
  cap_combo = c(lvc = log(41.7), lkel = log(0.0842), lk12 = log(0.0631), lk21 = log(0.0529))
)

# One arm = one solve, because rxSolve applies `params` globally and each arm
# needs a different PK parameter set. `id` and `arm` are assigned explicitly
# rather than via `keep`, since rxSolve drops `id` for a single-subject solve.
run_arm <- function(mod, pars, mg_per_kg, arm, id, tmax = 300, by = 1) {
  ev <- rbind(
    data.frame(time = 0, amt = mg_per_kg * WT * 1e6, dur = 30, evid = 1L,
               cmt = "central", dvid = NA_integer_),
    data.frame(time = seq(0, tmax, by = by), amt = NA_real_, dur = NA_real_,
               evid = 0L, cmt = "central", dvid = 1L)
  )
  out <- rxode2::rxSolve(mod, ev, params = pars, returnType = "data.frame",
                         useLinCmt = FALSE, atol = 1e-12, rtol = 1e-12)
  out$id <- id
  out$arm <- arm
  out$mg_per_kg <- mg_per_kg
  out$amt_ng <- mg_per_kg * WT * 1e6
  out
}
```

# Simulation

```{r simulate}
sims <- dplyr::bind_rows(
  # Figure 2 solid curves: monotherapy fits.
  run_arm(nif, pk_sets$nif_mono,  1.0, "Nifedipine 1.0 mg/kg (mono)",   1L),
  run_arm(cap, pk_sets$cap_mono, 15.0, "Captopril 15 mg/kg (mono)",     2L),
  # Figure 2 bold-solid curves: coadministration fits.
  run_arm(nif, pk_sets$nif_combo, 0.5, "Nifedipine 0.5 mg/kg (combo)",  3L),
  run_arm(cap, pk_sets$cap_combo, 5.0, "Captopril 5.0 mg/kg (combo)",   4L),
  # Figure 2 dashed curves: reduced dose predicted from the MONOTHERAPY PK.
  run_arm(nif, pk_sets$nif_mono,  0.5, "Nifedipine 0.5 mg/kg (predicted from mono PK)", 5L),
  run_arm(cap, pk_sets$cap_mono,  5.0, "Captopril 5.0 mg/kg (predicted from mono PK)",  6L)
)

drug_of <- function(a) ifelse(grepl("^Nifedipine", a), "Nifedipine", "Captopril")
sims$drug <- drug_of(sims$arm)
```

## Dose linearity

The paper states the PK "were nearly linear within these dosage ranges", which is
what licenses reusing the monotherapy PD parameters for the reduced
coadministration doses. With a fixed parameter set the model is exactly linear in
dose, so halving the nifedipine dose must halve every concentration. This is a
free, exact check on the infusion encoding.

```{r dose-linearity-gate}
lin <- sims |>
  dplyr::filter(arm %in% c("Nifedipine 1.0 mg/kg (mono)",
                           "Nifedipine 0.5 mg/kg (predicted from mono PK)"),
                time > 0) |>
  dplyr::select(time, arm, Cc) |>
  tidyr::pivot_wider(names_from = arm, values_from = Cc)

ratio <- lin[["Nifedipine 0.5 mg/kg (predicted from mono PK)"]] /
  lin[["Nifedipine 1.0 mg/kg (mono)"]]
stopifnot(max(abs(ratio - 0.5)) < 1e-8)
range(ratio)
```

# Replicate published figures

## Figure 2 -- plasma concentrations

```{r figure-2}
# Replicates Figure 2 of Kiriyama 2024: fitted plasma nifedipine (A) and
# captopril (B) after a 30-min IV infusion, monotherapy vs coadministration,
# plus the reduced dose predicted from the monotherapy PK (dashed in the paper).
sims |>
  ggplot(aes(time, Cc, colour = arm, linetype = arm)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~drug, scales = "free_y") +
  scale_x_continuous(breaks = seq(0, 300, by = 60)) +
  guides(colour = guide_legend(ncol = 1), linetype = guide_legend(ncol = 1)) +
  theme(legend.position = "bottom", legend.title = ggplot2::element_blank()) +
  labs(x = "Time (min)", y = "Plasma concentration (ng/mL)",
       title = "Figure 2 - fitted plasma concentrations",
       caption = "Replicates Figure 2 of Kiriyama 2024.")
```

```{r figure-2-gate}
peak30 <- sims |>
  dplyr::filter(time == 30) |>
  dplyr::select(arm, Cc) |>
  dplyr::mutate(Cc = round(Cc))
knitr::kable(
  peak30 |> dplyr::rename("Arm" = arm, "Cc at 30 min (ng/mL)" = Cc),
  caption = "End-of-infusion concentrations. Figure 2 read-offs: 2900 and 19500 ng/mL for the monotherapy solid curves, ~1200 and ~8300 ng/mL for the coadministration bold-solid curves."
)
pk_at <- function(a) peak30$Cc[peak30$arm == a]
stopifnot(
  abs(pk_at("Nifedipine 1.0 mg/kg (mono)")  - 2933)  <= 1,
  abs(pk_at("Captopril 15 mg/kg (mono)")    - 19591) <= 1,
  abs(pk_at("Nifedipine 0.5 mg/kg (combo)") - 1227)  <= 1,
  abs(pk_at("Captopril 5.0 mg/kg (combo)")  - 8492)  <= 1
)
```

## Figure 3 -- pharmacodynamics during monotherapy

```{r figure-3}
# Replicates Figure 3 of Kiriyama 2024: BP (A), HR (B), QT (C) and QTc (D)
# after nifedipine 1.0 mg/kg or captopril 15 mg/kg. QTc was only quantifiable
# for nifedipine (see Errata).
pd_long <- sims |>
  dplyr::filter(arm %in% c("Nifedipine 1.0 mg/kg (mono)", "Captopril 15 mg/kg (mono)"),
                time <= 180) |>
  dplyr::select(time, drug, bp, hr, qt, dplyr::any_of("qtc")) |>
  tidyr::pivot_longer(c(bp, hr, qt, dplyr::any_of("qtc")),
                      names_to = "endpoint", values_to = "value") |>
  dplyr::filter(!is.na(value)) |>
  dplyr::mutate(endpoint = factor(
    endpoint, levels = c("bp", "hr", "qt", "qtc"),
    labels = c("A: BP (mmHg)", "B: HR (beat/min)",
               "C: QT (msec)", "D: QTc (msec)")
  ))

pd_long |>
  ggplot(aes(time, value, colour = drug)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~endpoint, scales = "free_y") +
  scale_x_continuous(breaks = seq(0, 180, by = 60)) +
  theme(legend.position = "bottom", legend.title = ggplot2::element_blank()) +
  labs(x = "Time (min)", y = NULL,
       title = "Figure 3 - PD responses during monotherapy",
       caption = "Replicates Figure 3 of Kiriyama 2024. QTc (panel D) is nifedipine only.")
```

### Gates against the paper's own reported PD values

```{r figure-3-gates}
mono <- sims |> dplyr::filter(time <= 180)
nifm <- mono |> dplyr::filter(arm == "Nifedipine 1.0 mg/kg (mono)")
capm <- mono |> dplyr::filter(arm == "Captopril 15 mg/kg (mono)")

pd_checks <- tibble::tibble(
  Quantity = c("Nifedipine BP nadir (mmHg)",
               "Nifedipine BP nadir time (min)",
               "Nifedipine BP at 180 min (mmHg)",
               "Nifedipine HR at 180 min (beat/min)",
               "Captopril BP nadir (mmHg)",
               "Captopril HR at 180 min (beat/min)",
               "Captopril QT at 180 min (msec)"),
  Simulated = c(round(min(nifm$bp), 1),
                nifm$time[which.min(nifm$bp)],
                round(nifm$bp[nifm$time == 180], 1),
                round(nifm$hr[nifm$time == 180], 1),
                round(min(capm$bp), 1),
                round(capm$hr[capm$time == 180], 1),
                round(capm$qt[capm$time == 180], 1)),
  Paper = c("70.9 +/- 8.6 (measured Emax, Results)",
            "~10 (Results: 'approximately 10 min')",
            "~85-88 (Figure 3A)",
            "~250 (Results: 'approximately 250 beats/min')",
            "~80 at 40-60 min (Figure 3A)",
            "~250 (Results)",
            "~95-98 (Figure 3C)")
)
knitr::kable(pd_checks, caption = "Simulated PD landmarks against the values Kiriyama 2024 reports in its Results text and draws in Figure 3.")

stopifnot(
  # Nifedipine BP nadir sits inside the paper's measured Emax 70.9 +/- 8.6 mmHg,
  # and is pinned exactly as a regression guard.
  abs(min(nifm$bp) - 71.7) < 0.05,
  min(nifm$bp) > 70.9 - 8.6, min(nifm$bp) < 70.9 + 8.6,
  nifm$time[which.min(nifm$bp)] == 15,
  # BP recovers but never back to the 96.8 mmHg baseline (Figure 3A).
  nifm$bp[nifm$time == 180] > 80, nifm$bp[nifm$time == 180] < 90,
  max(nifm$bp) <= 96.8 + 1e-8,
  # The hypotensive arm's asymptote is a BP LEVEL of 67.1 mmHg, so BP can never
  # fall below it. Reading Emax,1 as a 67.1 mmHg decrease would allow ~29.7 mmHg.
  min(nifm$bp) > 67.1,
  # Both drugs converge on ~250 beat/min after 60-90 min (Results).
  abs(nifm$hr[nifm$time == 180] - 250) < 10,
  abs(capm$hr[capm$time == 180] - 250) < 10,
  # Captopril HR cannot fall below its own floor E0 - Emax = 232.4 beat/min.
  min(capm$hr) > 312.8 - 80.4,
  # Captopril QT plateau (Figure 3C).
  abs(capm$qt[capm$time == 180] - 96.5) < 2
)
```

### Drug-free gate on the BP homeostasis model

The nifedipine BP model is the only structure in the paper that required
reinterpretation (see Errata). The decisive check is that with **no drug at all**
the model must hold BP exactly at its baseline `E0`, which is precisely what the
paper's stated initial condition (`C3 = 0`, `C4 = 0`, `E = E0`) asserts. This is a
runtime gate rather than a comment because it is the single assertion that
distinguishes the correct reading of the EHS production term from the incorrect
one.

```{r drug-free-gate}
ev0 <- rbind(
  data.frame(time = 0, amt = 0, dur = 30, evid = 1L, cmt = "central", dvid = NA_integer_),
  data.frame(time = seq(0, 180, by = 1), amt = NA_real_, dur = NA_real_,
             evid = 0L, cmt = "central", dvid = 1L)
)
s0 <- rxode2::rxSolve(nif, ev0, params = pk_sets$nif_mono, returnType = "data.frame",
                      useLinCmt = FALSE, atol = 1e-12, rtol = 1e-12)

c(max_bp_deviation_mmHg = max(abs(s0$bp - 96.8)), max_ehs = max(abs(s0$ehs)))
stopifnot(max(abs(s0$bp - 96.8)) < 1e-6, max(abs(s0$ehs)) < 1e-6)
```

## Figure 4 -- pharmacodynamics during coadministration

The paper predicts the combination response by superposing each drug's separately
fitted effect (Methods 2.6):

* BP and HR: `Emix = E0 - ((E0 - E_nifedipine) + (E0 - E_captopril))`
* QT: `Emix = E0 + ((E_nifedipine - E0) + (E_captopril - E0))`

Both forms reduce to summing each drug's deviation from its own baseline and
adding it back onto a common baseline. Because the two model files carry
*different* fitted baselines for the same endpoint (BP: 96.8 vs 105 mmHg), the
common baseline has to be chosen; the nifedipine baseline is used here and the
choice is recorded in the Errata.

```{r figure-4}
# Replicates Figure 4 of Kiriyama 2024: BP (A), HR (B) and QT (C) during
# coadministration of nifedipine 0.5 mg/kg and captopril 5.0 mg/kg, using the
# coadministration PK of Table 2 and the monotherapy PD of Table 3, together
# with the paper's additive prediction (bold solid line in the paper).
nifc <- sims |> dplyr::filter(arm == "Nifedipine 0.5 mg/kg (combo)", time <= 180)
capc <- sims |> dplyr::filter(arm == "Captopril 5.0 mg/kg (combo)",  time <= 180)

e0_nif <- c(bp = 96.8, hr = 306,   qt = 75.8)
e0_cap <- c(bp = 105,  hr = 312.8, qt = 81.2)

additive <- tibble::tibble(
  time = nifc$time,
  bp = (nifc$bp - e0_nif[["bp"]]) + (capc$bp - e0_cap[["bp"]]) + e0_nif[["bp"]],
  hr = (nifc$hr - e0_nif[["hr"]]) + (capc$hr - e0_cap[["hr"]]) + e0_nif[["hr"]],
  qt = (nifc$qt - e0_nif[["qt"]]) + (capc$qt - e0_cap[["qt"]]) + e0_nif[["qt"]]
)

combo_long <- dplyr::bind_rows(
  nifc     |> dplyr::select(time, bp, hr, qt) |> dplyr::mutate(series = "Nifedipine 0.5 mg/kg"),
  capc     |> dplyr::select(time, bp, hr, qt) |> dplyr::mutate(series = "Captopril 5.0 mg/kg"),
  additive |> dplyr::mutate(series = "Additive prediction")
) |>
  tidyr::pivot_longer(c(bp, hr, qt), names_to = "endpoint", values_to = "value") |>
  dplyr::mutate(endpoint = factor(endpoint, levels = c("bp", "hr", "qt"),
                                  labels = c("A: BP (mmHg)", "B: HR (beat/min)",
                                             "C: QT (msec)")))

combo_long |>
  ggplot(aes(time, value, colour = series, linewidth = series)) +
  geom_line() +
  facet_wrap(~endpoint, scales = "free_y") +
  scale_linewidth_manual(values = c("Additive prediction" = 1.1,
                                    "Nifedipine 0.5 mg/kg" = 0.6,
                                    "Captopril 5.0 mg/kg" = 0.6)) +
  scale_x_continuous(breaks = seq(0, 180, by = 60)) +
  theme(legend.position = "bottom", legend.title = ggplot2::element_blank()) +
  labs(x = "Time (min)", y = NULL,
       title = "Figure 4 - PD responses during coadministration",
       caption = "Replicates Figure 4 of Kiriyama 2024; the bold line is the paper's additive prediction.")
```

The paper's qualitative conclusions about the combination are reproduced: the
early BP fall is dominated by nifedipine, captopril contributes little to HR at
5.0 mg/kg, and the additive prediction sits below each single agent for QT.

```{r figure-4-gate}
combo_checks <- tibble::tibble(
  Quantity = c("BP nadir, nifedipine arm (mmHg)",
               "BP nadir, additive prediction (mmHg)",
               "HR change from baseline, captopril arm at 180 min (beat/min)",
               "QT at 180 min, additive prediction (msec)"),
  Simulated = c(round(min(nifc$bp), 1), round(min(additive$bp), 1),
                round(capc$hr[capc$time == 180] - e0_cap[["hr"]], 1),
                round(additive$qt[additive$time == 180], 1))
)
knitr::kable(combo_checks, caption = "Coadministration landmarks. The paper reports a measured combination BP Emax of 75.0 +/- 4.8 mmHg approximately 10 min after infusion start (Results, Figure 4A).")

stopifnot(
  # The paper: "the alteration in BP during coadministration was nearly equal to
  # that produced by nifedipine", with measured Emax 75.0 +/- 4.8 mmHg.
  min(nifc$bp) > 75.0 - 4.8, min(nifc$bp) < 75.0 + 4.8,
  # Captopril's own HR contribution at 5.0 mg/kg is modest ("Captopril exhibited
  # minimal effects on HR at this dose") but not zero.
  capc$hr[capc$time == 180] - e0_cap[["hr"]] < 0,
  # The additive prediction must be more extreme than either single agent for BP.
  min(additive$bp) < min(nifc$bp)
)
```

# PKNCA validation

Table 1 of the paper reports moment-analysis NCA for all four dose arms. The NCA
below runs on the model-predicted profiles over the paper's own 5-300 min
sampling window, using the PK parameter set matching each arm.

```{r pknca}
nca_arms <- c("Nifedipine 1.0 mg/kg (mono)", "Nifedipine 0.5 mg/kg (combo)",
              "Captopril 15 mg/kg (mono)",   "Captopril 5.0 mg/kg (combo)")

sim_nca <- sims |>
  dplyr::filter(arm %in% nca_arms, !is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

# Guarantee a time = 0 row per arm; for a 30-min infusion the pre-dose
# concentration is 0.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, arm, time)

dose_df <- sims |>
  dplyr::filter(arm %in% nca_arms) |>
  dplyr::distinct(id, arm, amt_ng) |>
  dplyr::mutate(time = 0, amt = amt_ng, dur = 30) |>
  dplyr::select(id, time, amt, dur, arm)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id,
                             concu = "ng/mL", timeu = "min")
# `duration` is mandatory for an infusion: without it PKNCA treats the dose as a
# bolus and the IV MRT / Vss are inflated by dur/2.
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id,
                             route = "intravascular", duration = "dur",
                             doseu = "ng")

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE,
  cl.obs = TRUE, vz.obs = TRUE, mrt.iv.obs = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

## Comparison against published NCA (Table 1)

Two transcription notes. First, Table 1 reports AUC in ug*min/mL; the values
below are multiplied by 1000 so both sides are in ng*min/mL. Second, the paper
states that "the volume of distribution at steady state (Vdss) was calculated
using CLtot/ke" -- that is CL divided by the terminal rate constant, which is
**Vz**, not Vss. The Table 1 "Vdss" row is therefore compared against PKNCA's
`vz.obs`.

```{r nca-compare}
published <- tibble::tribble(
  ~arm,                            ~aucinf.obs, ~half.life, ~cl.obs, ~vz.obs, ~mrt.iv.obs,
  "Nifedipine 1.0 mg/kg (mono)",      237e3,        69.0,     1.35,     162,        82.8,
  "Nifedipine 0.5 mg/kg (combo)",      63.4e3,      53.8,     2.24,     161,        39.0,
  "Captopril 15 mg/kg (mono)",       1730e3,        81.6,     4.21,     395,        69.8,
  "Captopril 5.0 mg/kg (combo)",      411e3,       190.0,     4.92,    1565,        98.9
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by        = "arm",
  units     = c(aucinf.obs = "ng*min/mL", half.life = "min", cl.obs = "mL/min",
                vz.obs = "mL", mrt.iv.obs = "min"),
  tolerance_pct = 20
)

knitr::kable(cmp, caption = "Simulated vs Kiriyama 2024 Table 1. * differs from the reference by >20%.")
attr(cmp, "footnote")
```

Ten of the twenty rows are starred. The reason is methodological rather than a
transcription error, but it is worth being specific about which explanation
actually holds.

Table 1 is a **moment analysis of individual animals, then averaged** (3-6 rats
per arm, i.e. `mean(D / AUC_i)`), whereas Table 2 -- the parameter set these
model files encode -- is a **naive-pooled fit to the mean profile**
(`D / AUC_mean`). Those two estimators diverge whenever the between-animal
spread is wide, and here it is extreme: several of Table 1's own standard
deviations equal or exceed their means (`CLtot` 4.21 +/- 4.25 mL/min for
captopril monotherapy, `Vdss` 1565 +/- 1695 mL and `t1/2` 190 +/- 157 min for
captopril combination). The paper itself flags a mismatch between its two tables
-- "the Cmax of nifedipine was overestimated during monotherapy, whereas for
captopril, it was underestimated".

One candidate explanation was tested and **rejected**: the paper estimates
`t1/2` by "linear regression of three minimal data points from the terminal
portion". Recomputing the simulated `t1/2` that way -- restricting the profile to
the Figure 2 sampling times and regressing only the last three -- reproduces
PKNCA's default almost exactly (nifedipine monotherapy 79.5 vs 79.1 min;
captopril monotherapy 50.0 vs 49.8 min), so the terminal-slope rule is *not*
what separates the two tables. That explanation is therefore not offered.

The informative comparison for this paper is containment within Table 1's own
reported variability, since a 20% tolerance is far tighter than the data support:

```{r nca-sd-containment}
published_sd <- tibble::tribble(
  ~arm,   ~param,        ~mean,   ~sd,
  nca_arms[1], "aucinf.obs",  237e3,   85e3,
  nca_arms[2], "aucinf.obs",   63.4e3, 23.4e3,
  nca_arms[3], "aucinf.obs", 1730e3,  411e3,
  nca_arms[4], "aucinf.obs",  411e3,  407e3,
  nca_arms[1], "half.life",    69.0,   41.9,
  nca_arms[2], "half.life",    53.8,   17.1,
  nca_arms[3], "half.life",    81.6,   58.4,
  nca_arms[4], "half.life",   190.0,  157.0,
  nca_arms[1], "cl.obs",        1.35,   0.57,
  nca_arms[2], "cl.obs",        2.24,   1.30,
  nca_arms[3], "cl.obs",        4.21,   4.25,
  nca_arms[4], "cl.obs",        4.92,   2.57,
  nca_arms[1], "vz.obs",      162,    171,
  nca_arms[2], "vz.obs",      161,     61,
  nca_arms[3], "vz.obs",      395,    334,
  nca_arms[4], "vz.obs",     1565,   1695,
  nca_arms[1], "mrt.iv.obs",   82.8,   60.8,
  nca_arms[2], "mrt.iv.obs",   39.0,    3.6,
  nca_arms[3], "mrt.iv.obs",   69.8,   26.7,
  nca_arms[4], "mrt.iv.obs",   98.9,   70.5
)

sim_vals <- as.data.frame(nca_res$result) |>
  dplyr::filter(start == 0, is.infinite(end)) |>
  dplyr::select(arm, param = PPTESTCD, simulated = PPORRES)

containment <- published_sd |>
  dplyr::left_join(sim_vals, by = c("arm", "param")) |>
  dplyr::mutate(
    Parameter = nlmixr2lib::ncaParamLabel(param),
    `Table 1 mean +/- SD` = paste0(signif(mean, 3), " +/- ", signif(sd, 3)),
    Simulated = signif(simulated, 3),
    `Within 1 SD` = ifelse(simulated >= mean - sd & simulated <= mean + sd, "yes", "NO")
  )

knitr::kable(
  containment |> dplyr::select(Parameter, arm, `Table 1 mean +/- SD`, Simulated, `Within 1 SD`) |>
    dplyr::rename("Arm" = arm),
  caption = "Simulated NCA against Table 1's own mean +/- 1 SD."
)

n_within <- sum(containment$`Within 1 SD` == "yes")
c(within_1_SD = n_within, total = nrow(containment))
stopifnot(n_within >= 16)
```

Sixteen of the twenty simulated values fall inside Table 1's mean +/- 1 SD, and
the four that do not form a coherent pattern rather than scattering: every one is
a **terminal-phase** quantity (`t1/2`, the `CLtot/ke` volume derived from it, or
MRT) in a **reduced-dose coadministration** arm. That is expected from Table 2
itself, whose combination fits carry much faster micro-constants than the
monotherapy fits (nifedipine `k10` rises from 0.0218 to 0.0768 min<sup>-1</sup>)
and therefore imply a far shorter terminal phase than the moment analysis of the
same animals produced. It is an internal inconsistency between the paper's
Table 1 and Table 2 for the combination arms, not a defect of the encoding -- and
notably the paper reports *better* AIC for both combination fits. No parameter
was tuned.

The consistency check that *is* meaningful here is against Table 2's own fitted
quantities, which the model must reproduce exactly:

```{r nca-table2-gate}
kel_vc <- tibble::tibble(
  Arm = nca_arms,
  `CL = k10 * V1 (mL/min)` = round(c(0.0218 * 54.9, 0.0768 * 26.7,
                                     0.0521 * 54.2, 0.0842 * 41.7), 3)
)
knitr::kable(kel_vc, caption = "Clearance implied by the Table 2 micro-constants, CL = k10 * V1.")

cl_sim <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD == "cl.obs", start == 0, is.infinite(end))
cl_expected <- c(0.0218 * 54.9, 0.0768 * 26.7, 0.0521 * 54.2, 0.0842 * 41.7)
names(cl_expected) <- nca_arms
# Model-based CL from the NCA must equal k10 * V1 to within the numerical
# tolerance of the AUC extrapolation.
stopifnot(
  max(abs(cl_sim$PPORRES - cl_expected[cl_sim$arm]) / cl_expected[cl_sim$arm]) < 0.01
)
round(setNames(cl_sim$PPORRES, cl_sim$arm), 4)
```

# Assumptions and deviations

* **EHS production term (the one structural reinterpretation).** Methods 2.5
  prints the endogenous-hypertensive-substance ODE as
  `dC4/dt = K_EHS,in * (1 + (E0 - E)/E0 * alpha) - k_EHS,out * C4`
  and, three lines later, the initial condition
  `at t = 0, C3 = 0, C4 = 0, and E = E0`. Those two statements are only mutually
  consistent if `C4` is the EHS **deviation from its own drug-free steady state**:
  the printed ODE is the total-EHS equation, whose drug-free steady state is
  `K_EHS,in / k_EHS,out = 1.824 units`, and substituting
  `delta = C4_total - K_EHS,in/k_EHS,out` cancels the constant term exactly,
  giving `d(delta)/dt = K_EHS,in * alpha * (E0 - E)/E0 - k_EHS,out * delta` with
  `delta(0) = 0`. The model files encode that deviation form. Keeping the
  `1 +` while also starting from `C4 = 0` double-counts baseline production and
  is refuted three independent ways: BP would drift to 131 mmHg with **no drug
  present** (violating the paper's own initial condition), the nifedipine BP
  nadir would be 78.5 mmHg against a measured 70.9 +/- 8.6 mmHg, and BP at
  180 min would be 118.9 mmHg -- 22 mmHg above the untreated baseline, where
  Figure 3A shows recovery only to about 85-88 mmHg. The deviation form
  reproduces all three (drug-free BP held at 96.8 mmHg, nadir 71.7 mmHg at
  15 min, 84.3 mmHg at 180 min); the drug-free case is asserted above as a
  runtime gate. No parameter value was altered.
* **Emax in the nifedipine BP model is an asymptotic level, not a change.** The
  printed `dE/dt` carries the factors `(E0 - Emax,1)` and `(Emax,2 - E0)`, so
  integrating it makes each arm interpolate from `E0` toward its `Emax`. Reading
  `Emax,1 = 67.1 mmHg` as a decrease instead would put the nadir near 29.7 mmHg,
  which Figure 3A refutes; as an asymptote it gives 71.7 mmHg against the
  measured 70.9 +/- 8.6 mmHg. The captopril model and the nifedipine HR / QT /
  QTc models use the ordinary form where `Emax` *is* a change from baseline.
* **Body weight (dose scale).** Not stated beyond 215-315 g. Back-solved as
  ~250 g from the Figure 2 fitted peaks; nifedipine gives 247 g and captopril
  249 g independently, and the agreement is asserted as a gate. The weight is a
  simulation input only -- it is not a model covariate, because the paper fits
  absolute volumes and rate constants.
* **`alpha` units.** Table 3 prints `alpha` as mmHg/min, but the equation
  multiplies the dimensionless ratio `(E0 - E)/E0`, so `alpha` must be
  dimensionless. The printed unit is treated as a typographical slip; the value
  1.17 is used as printed.
* **QTc parameters come from the Results prose, not a table.** The five
  nifedipine QTc values (`ke0 = 0.0130`, `EC50 = 522`, `E0 = 169`,
  `Emax = 29.5`, `gamma = 1.92`) appear only in the Results narrative
  accompanying Figure 3D. Captopril has no QTc sub-model at all: the paper
  states the relationship "could not be obtained" because of the transient HR
  rise, so `Kiriyama_2024_captopril_rat` carries three endpoints and
  `Kiriyama_2024_nifedipine_rat` four.
* **The transient HR rise after captopril is not in the model.** The paper
  measures an HR `Emax` of 325 +/- 35 beat/min against a 312.8 beat/min baseline
  immediately after captopril infusion, but its sigmoid `Emax` model is purely
  inhibitory and captures only the subsequent reduction. The extraction is
  faithful to the fitted model, so simulated captopril HR is monotonically
  decreasing and will not show the early spike visible in Figures 3B and 4B.
* **Coadministration PK is applied as a scenario, not a covariate.** Table 2
  gives separate PK parameters for each drug under coadministration, and they
  differ substantially (nifedipine `V1` 54.9 -> 26.7 mL). The paper concludes
  that "no significant PK interactions were observed" and that none of the
  Table 1 differences reached significance, so encoding an interaction covariate
  would assert an effect the authors declined to claim. The `ini()` block holds
  the monotherapy values and the vignette applies the coadministration set via
  `rxSolve(params = )`.
* **Additive-prediction baseline.** The paper writes a single `E0` in its
  superposition formulas, but the two drugs have separately fitted baselines for
  the same endpoint (BP 96.8 vs 105 mmHg; the paper notes the group baselines
  did not differ significantly). Figure 4 above sums each drug's deviation from
  its own fitted baseline and adds the result to the nifedipine baseline.
* **No IIV and no residual error.** Both fits are naive-pooled WinNonlin fits to
  mean profiles and report only AIC. No `eta` terms are present and every
  residual SD is `fixed(0)` rather than invented, so the models reproduce the
  published typical-value curves exactly and are deterministic.
* **Supplement not required.** Figures S1-S4 hold concentration-versus-effect
  plots and the parameter sensitivity checks; they contain no parameter values
  beyond Tables 1-3 and the Results prose, so no supplement was needed.
* **Anaesthesia confounder (paper's own caveat).** Urethane inhibits CYP3A4 and
  nifedipine is a CYP3A4 substrate, so the Discussion flags a possible effect on
  nifedipine PK. Baseline BP under anaesthesia is also lower than in awake SHRs
  (the paper reports 97.8 mmHg anaesthetised). These values are extracted as
  published; they describe anaesthetised animals.
