---
title: "Vancomycin (Kirwan 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vancomycin (Kirwan 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta}
mod <- readModelDb("Kirwan_2021_vancomycin")
ui <- rxode2::rxode(mod)
```

* Citation: `r ui$reference`
* Description: `r ui$description`
* Article: <https://doi.org/10.1186/s13054-021-03863-4> (open access, CC BY 4.0)
* Supplement (Additional file 1, Tables S1-S2 and Figures S1-S2): available from
  the article page.

Kirwan and colleagues fitted a one-compartment model to routine peak and trough
therapeutic-drug-monitoring (TDM) levels with the non-parametric adaptive grid
(NPAG) algorithm in Pmetrics. They screened a long list of continuous and
categorical covariates, found that none gave a clear improvement, and used the
**base structural model** -- clearance and volume only -- for their
probability-of-target-attainment (PTA) dosing simulations. That base model is
the model packaged here.

# Population

The analysis used 106 vancomycin dosing intervals from 24 critically ill adults
in a single tertiary intensive care unit in Dublin, Ireland, all receiving
continuous venovenous haemodiafiltration (CVVHDF) for most of each included
dosing interval (Methods; Table 1). Eighteen of 24 (75%) were male; mean age
was 65.5 years (SD 12.3), mean weight 81.8 kg (SD 24), mean BMI 28.6 kg/m^2.
Patients were severely ill (APACHE II mean 23, SOFA mean 10.5, in-hospital
mortality 25%) and largely anuric (median urine output on study day 1,
77.9 mL/24 h). The ICU dosing policy was a 25 mg/kg loading dose followed by
15-20 mg/kg once daily, infused at 10 mg/min; the doses actually recorded
averaged 1098 mg (SD 249) at a mean interval of 18.5 h (Table 2). Peaks were
drawn 60 min after the end of the infusion and troughs immediately before the
next dose.

```{r population}
str(ui$population[c("n_subjects", "n_dosing_intervals", "sex_female_pct",
                    "age_range", "weight_range", "regions")])
```

# Source trace

Every value below is also carried as an in-file comment next to its `ini()`
entry in `inst/modeldb/specificDrugs/Kirwan_2021_vancomycin.R`.

| Model element | Value | Source location |
|---|---|---|
| Structural model | 1-compartment, IV infusion, first-order elimination | Methods, 'Population pharmacokinetic analysis' |
| `lcl` (total CL on CVVHDF) | log(2.59 L/h) | Table 3, base model CL mean (SD 0.49, CV 18.99%, median 2.70); Results text |
| `lvc` (V) | log(80.98 L) | Table 3, base model V mean (SD 16.89, CV 20.86%, median 73.72); Results text |
| `etalcl` | 0.035427 = log(0.1899^2 + 1) | Table 3, base model CL CV% |
| `etalvc` | 0.042594 = log(0.2086^2 + 1) | Table 3, base model V CV% |
| `addSd`, `propSd` | fixed(0) | Not reported (see Assumptions and deviations) |
| `d/dt(central) <- -kel * central`, `Cc <- central / vc` | n/a | Methods, 'Population pharmacokinetic analysis' |
| Infusion rate 10 mg/min | n/a | Methods, 'Vancomycin therapy and TDM' |
| PTA regimens 1-4 | n/a | Methods, 'Probability of target attainment (PTA) plots'; Figure 2; Additional file 1 Figures S1-S2 |

# Virtual cohort

Original observed data are not publicly available. The final model has no
covariates, so a virtual subject is defined by its dosing regimen alone. The
cohort simulates the four exploratory regimens of the paper's PTA analysis
(Methods), with 200 virtual patients per regimen (the paper used 1000):

1. 2 g loading dose, then 750 mg every 12 h from 12 h;
2. 2 g loading dose, then 500 mg every 12 h from 12 h;
3. 1.5 g loading dose, then 750 mg every 12 h from 12 h;
4. 2 g loading dose, then 1.5 g at 12 h and 1.5 g every 24 h thereafter.

Every dose is infused at the ICU's 10 mg/min (600 mg/h). Observations run on a
0.25-h grid to 48 h on the `central` state; because the doses are infusions,
the concentration at a dose time equals the pre-dose (trough) value.

```{r cohort}
regimens <- list(
  "R1: 2 g then 750 mg q12h" = data.frame(time = c(0, 12, 24, 36), amt = c(2000, 750, 750, 750)),
  "R2: 2 g then 500 mg q12h" = data.frame(time = c(0, 12, 24, 36), amt = c(2000, 500, 500, 500)),
  "R3: 1.5 g then 750 mg q12h" = data.frame(time = c(0, 12, 24, 36), amt = c(1500, 750, 750, 750)),
  "R4: 2 g, 1.5 g at 12 h, then q24h" = data.frame(time = c(0, 12, 36), amt = c(2000, 1500, 1500))
)
n_per_regimen <- 200L
obs_times <- seq(0, 48, by = 0.25)

make_cohort <- function(regimen, doses, n, id_offset) {
  ids <- id_offset + seq_len(n)
  dose_rows <- tidyr::expand_grid(id = ids, doses) |>
    dplyr::mutate(evid = 1L, cmt = "central", rate = 600)
  obs_rows <- tidyr::expand_grid(id = ids, time = obs_times) |>
    dplyr::mutate(evid = 0L, amt = NA_real_, cmt = "central", rate = NA_real_)
  dplyr::bind_rows(dose_rows, obs_rows) |>
    dplyr::mutate(regimen = regimen) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(lapply(seq_along(regimens), function(i) {
  make_cohort(names(regimens)[i], regimens[[i]], n_per_regimen,
              id_offset = (i - 1L) * n_per_regimen)
}))
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
rxode2::rxSetSeed(20211210)
sim <- rxode2::rxSolve(mod, events = events, keep = "regimen") |>
  as.data.frame()
```

## Typical-value check against the closed form

With the random effects zeroed, the one-compartment infusion model has a
closed-form superposition solution. Both sides use the same parameters, so the
difference is pure numerical error and a tight bound is appropriate.

```{r typical-check}
sim_typ <- rxode2::rxSolve(rxode2::zeroRe(mod),
                           events = dplyr::filter(events, id == 1L)) |>
  as.data.frame()

cl <- 2.59; v <- 80.98; k <- cl / v
closed_form <- function(t, doses, rate = 600) {
  vapply(t, function(tt) {
    sum(vapply(seq_len(nrow(doses)), function(j) {
      dur <- doses$amt[j] / rate
      ts <- tt - doses$time[j]
      if (ts <= 0) return(0)
      r0 <- rate / cl
      if (ts <= dur) r0 * (1 - exp(-k * ts))
      else r0 * (1 - exp(-k * dur)) * exp(-k * (ts - dur))
    }, numeric(1)))
  }, numeric(1))
}
cf <- closed_form(sim_typ$time, regimens[[1]])
stopifnot(max(abs(sim_typ$Cc - cf) / pmax(cf, 1)) < 1e-4)
```

The typical patient on regimen 1 has a trough of
`r round(cf[sim_typ$time == 12], 1)` mg/L at 12 h and
`r round(cf[sim_typ$time == 48], 1)` mg/L at 48 h; the elimination half-life is
`r round(log(2) / k, 1)` h.

```{r profile-plot, fig.cap = "Simulated vancomycin concentration by regimen: median and 5th-95th percentiles over the first 48 h (200 virtual patients per regimen)."}
sim |>
  dplyr::group_by(regimen, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  geom_hline(yintercept = c(15, 20), linetype = "dashed", colour = "grey40") +
  facet_wrap(~regimen) +
  scale_x_continuous(breaks = seq(0, 48, 12)) +
  labs(x = "Time after first dose (h)", y = "Vancomycin Cc (mg/L)")
```

# Replicate Figure 2 (trough PTA)

Figure 2 of Kirwan 2021 plots, for each regimen, the proportion of simulated
patients whose pre-dose concentration at 12, 24, 36 and 48 h reaches 10, 15,
18, 20 and 25 mg/L. The published points below were digitised by the
maintainers from the high-resolution Figure 2 raster (reading resolution about
+/-0.02). At 12 h the three 2 g-loading regimens are the same regimen; the
black regimen-1 line is hidden under the other two and was not digitised.
Regimen 4 has no trough at 24 or 48 h.

```{r fig2-published}
fig2_pub <- tibble::tribble(
  ~regimen_short, ~time, ~conc, ~prop_pub,
  "R2", 12, 15, 0.86, "R2", 12, 18, 0.43, "R2", 12, 20, 0.17,
  "R4", 12, 15, 0.82, "R4", 12, 18, 0.37, "R4", 12, 20, 0.13,
  "R3", 12, 15, 0.11, "R3", 12, 18, 0.00, "R3", 12, 20, 0.00,
  "R1", 24, 15, 0.93, "R1", 24, 18, 0.49, "R1", 24, 20, 0.22,
  "R2", 24, 15, 0.67, "R2", 24, 18, 0.19, "R2", 24, 20, 0.06,
  "R3", 24, 15, 0.49, "R3", 24, 18, 0.08, "R3", 24, 20, 0.01,
  "R1", 36, 15, 0.96, "R1", 36, 18, 0.56, "R1", 36, 20, 0.29,
  "R2", 36, 15, 0.49, "R2", 36, 18, 0.11, "R2", 36, 20, 0.02,
  "R3", 36, 15, 0.73, "R3", 36, 18, 0.24, "R3", 36, 20, 0.08,
  "R4", 36, 15, 0.76, "R4", 36, 18, 0.30, "R4", 36, 20, 0.13,
  "R1", 48, 15, 0.97, "R1", 48, 18, 0.61, "R1", 48, 20, 0.34,
  "R2", 48, 15, 0.38, "R2", 48, 18, 0.08, "R2", 48, 20, 0.02,
  "R3", 48, 15, 0.86, "R3", 48, 18, 0.41, "R3", 48, 20, 0.17
)

pta_trough <- function(s) {
  s |>
    dplyr::filter(time %in% c(12, 24, 36, 48)) |>
    dplyr::mutate(regimen_short = substr(regimen, 1, 2)) |>
    dplyr::filter(!(regimen_short == "R4" & time %in% c(24, 48))) |>
    tidyr::crossing(conc = c(10, 15, 18, 20, 25)) |>
    dplyr::group_by(regimen, regimen_short, time, conc) |>
    dplyr::summarise(prop_sim = mean(Cc >= conc), .groups = "drop")
}
fig2_sim <- pta_trough(sim)
fig2_cmp <- dplyr::inner_join(fig2_sim, fig2_pub,
                              by = c("regimen_short", "time", "conc")) |>
  dplyr::mutate(diff_pts = 100 * (prop_sim - prop_pub))
```

```{r fig2-plot, fig.height = 6, fig.cap = "Replicates Figure 2 of Kirwan 2021: proportion of patients whose pre-dose concentration reaches each threshold. Lines = this model (200 virtual patients per regimen); points = digitised Figure 2."}
ggplot(fig2_sim, aes(conc, prop_sim, colour = regimen)) +
  geom_line() +
  geom_point(data = dplyr::left_join(fig2_pub,
                                     dplyr::distinct(fig2_sim, regimen, regimen_short),
                                     by = "regimen_short"),
             aes(y = prop_pub), size = 2) +
  facet_wrap(~time, labeller = labeller(time = function(x) paste0("Pre-dose at ", x, " h"))) +
  scale_x_continuous(breaks = c(10, 15, 18, 20, 25)) +
  labs(x = "Trough concentration threshold (mg/L)", y = "Proportion with success",
       colour = NULL) +
  theme(legend.position = "bottom", legend.direction = "vertical")
```

```{r fig2-table}
fig2_cmp |>
  dplyr::filter(conc == 15 | conc == 20) |>
  dplyr::mutate(prop_sim = round(prop_sim, 2), diff_pts = round(diff_pts)) |>
  dplyr::select(regimen, time, conc, prop_pub, prop_sim, diff_pts) |>
  dplyr::arrange(regimen, conc, time) |>
  dplyr::rename(
    "Regimen" = regimen, "Time (h)" = time, "Threshold (mg/L)" = conc,
    "Published (Fig. 2)" = prop_pub, "Simulated" = prop_sim,
    "Difference (percentage points)" = diff_pts
  ) |>
  knitr::kable(caption = "Probability of a pre-dose level at or above 15 and 20 mg/L.")
```

Agreement is within a few percentage points almost everywhere. The one
consistent gap is at the top of regimen 1, where the model reaches 15 mg/L in
a few percent fewer patients than the paper (about 0.90 against 0.93-0.97).
That is the expected signature of replacing the non-parametric NPAG density
with a log-normal of the same CV: the log-normal has a slightly longer lower
tail of high-clearance or large-volume patients.

The paper's Results text reads the same curves in words; for example regimen 3
reaches 15 mg/L in "10-20% at 12 h, 50% at 24 h, 70% at 36 h and > 80% at
48 h", and "the probability of a level of 25 mg/L was negligible at all
investigated time points for each of the simulated dosage regimens".

```{r fig2-gate}
p25 <- fig2_sim |> dplyr::filter(conc == 25)
stopifnot(
  # Centre: a mis-transcribed CL, V, dose or infusion rate moves these
  # proportions by tens of percentage points.
  median(abs(fig2_cmp$diff_pts)) < 8,
  # Envelope: robust quantile, not the maximum (each point carries ~3.5
  # percentage points of Monte Carlo error at 200 patients).
  quantile(abs(fig2_cmp$diff_pts), 0.9) < 18,
  # 'Negligible' at 25 mg/L (Results).
  median(p25$prop_sim) < 0.05
)
```

## Which column of Table 3 is the typical value?

Pmetrics reports both the mean and the median of each parameter's
non-parametric distribution, and for this model they differ (CL 2.59 vs
2.70 L/h, V 80.98 vs 73.72 L). The Results text quotes the means. Re-running the
same cohort with the medians shows that the means also reproduce the paper's own
PTA simulation more closely, which is why the means are encoded.

```{r mean-vs-median}
mod_median <- mod |> rxode2::ini(lcl = log(2.70), lvc = log(73.72))
rxode2::rxSetSeed(20211210)
sim_median <- rxode2::rxSolve(mod_median, events = events, keep = "regimen") |>
  as.data.frame()
fig2_median <- dplyr::inner_join(pta_trough(sim_median), fig2_pub,
                                 by = c("regimen_short", "time", "conc")) |>
  dplyr::mutate(diff_pts = 100 * (prop_sim - prop_pub))

tibble::tibble(
  "Typical values" = c("Table 3 means (encoded)", "Table 3 medians"),
  "Median |difference| (points)" = round(c(median(abs(fig2_cmp$diff_pts)),
                                           median(abs(fig2_median$diff_pts))), 1),
  "Mean |difference| (points)" = round(c(mean(abs(fig2_cmp$diff_pts)),
                                         mean(abs(fig2_median$diff_pts))), 1)
) |>
  knitr::kable(caption = "Agreement with digitised Figure 2 under each column of Table 3.")
```

# PKNCA validation: AUC over 24-48 h (Figure S1)

Additional file 1 Figure S1 gives, per regimen, the proportion of simulated
patients whose AUC from 24 to 48 h reaches 100-600 mg*h/L. The figure is a
vector graphic, so the maintainers read its plotted points exactly from the
figure's drawing coordinates.

```{r pknca, message = FALSE}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, regimen)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::mutate(duration = amt / rate) |>
  dplyr::select(id, time, amt, duration, regimen)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | regimen + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | regimen + id,
                             route = "intravascular", duration = "duration")
intervals <- data.frame(start = 24, end = 48, auclast = TRUE, cmax = TRUE, cmin = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

auc_ind <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(id, regimen, auc = PPORRES)
```

```{r figS1}
figS1_pub <- tibble::tribble(
  ~regimen_short, ~`100`, ~`200`, ~`300`, ~`400`, ~`500`, ~`600`,
  "R1", 1, 1, 1, 1.000, 0.660, 0.200,
  "R2", 1, 1, 1, 0.628, 0.111, 0.000,
  "R3", 1, 1, 1, 0.935, 0.390, 0.041,
  "R4", 1, 1, 1, 1.000, 0.843, 0.369
) |>
  tidyr::pivot_longer(-regimen_short, names_to = "auc_threshold", values_to = "prop_pub") |>
  dplyr::mutate(auc_threshold = as.numeric(auc_threshold))

figS1_cmp <- auc_ind |>
  dplyr::mutate(regimen_short = substr(regimen, 1, 2)) |>
  tidyr::crossing(auc_threshold = c(100, 200, 300, 400, 500, 600)) |>
  dplyr::group_by(regimen, regimen_short, auc_threshold) |>
  dplyr::summarise(prop_sim = mean(auc >= auc_threshold), .groups = "drop") |>
  dplyr::inner_join(figS1_pub, by = c("regimen_short", "auc_threshold"))

figS1_cmp |>
  dplyr::filter(auc_threshold >= 400) |>
  dplyr::mutate(prop_sim = round(prop_sim, 2)) |>
  dplyr::select(regimen, auc_threshold, prop_pub, prop_sim) |>
  dplyr::rename(
    "Regimen" = regimen, "AUC24-48 threshold (mg*h/L)" = auc_threshold,
    "Published (Fig. S1)" = prop_pub, "Simulated" = prop_sim
  ) |>
  knitr::kable(caption = "Probability of AUC24-48 at or above each threshold (all regimens reach 1.00 up to 300 mg*h/L in both).")
```

The paper's headline AUC claim is that regimen 1 gives "approximately a 100%
probability of attaining a target AUC of 400 mg/L * h for MIC 1 mg/L".

## Comparison against the published AUC24-48

The paper reports no NCA table. The median AUC24-48 of each regimen's simulated
population is, however, fixed by Figure S1: it is the AUC at which the published
curve crosses 0.5, obtained here by log-linear interpolation between the two
bracketing plotted points.

```{r nca-compare}
auc_median_pub <- figS1_pub |>
  dplyr::group_by(regimen_short) |>
  dplyr::arrange(auc_threshold, .by_group = TRUE) |>
  dplyr::summarise(
    auclast = {
      i <- max(which(prop_pub >= 0.5))
      x0 <- auc_threshold[i]; x1 <- auc_threshold[i + 1]
      p0 <- prop_pub[i]; p1 <- prop_pub[i + 1]
      exp(log(x0) + (p0 - 0.5) / (p0 - p1) * log(x1 / x0))
    },
    .groups = "drop"
  ) |>
  dplyr::left_join(dplyr::distinct(auc_ind, regimen) |>
                     dplyr::mutate(regimen_short = substr(regimen, 1, 2)),
                   by = "regimen_short") |>
  dplyr::select(regimen, auclast)

nca_sim_auc <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD == "auclast")

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_sim_auc,
  reference = auc_median_pub,
  by = "regimen",
  units = c(auclast = "mg*h/L"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Median AUC24-48: simulated vs. read from Figure S1. * differs from reference by >20%.")

auc_pct <- as.data.frame(nca_sim_auc) |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(sim = median(PPORRES), .groups = "drop") |>
  dplyr::inner_join(auc_median_pub, by = "regimen") |>
  dplyr::mutate(pct_diff = 100 * (sim / auclast - 1))
stopifnot(
  # Centre of each regimen's AUC distribution; ~2% Monte Carlo error on a
  # median of 200 patients.
  all(abs(auc_pct$pct_diff) < 10),
  # Regimen 1: 'approximately a 100% probability' of AUC24-48 >= 400.
  mean(auc_ind$auc[startsWith(auc_ind$regimen, "R1")] >= 400) > 0.9
)
```

# Replicate Figure S2 (peak levels after the 36 h dose)

Figure S2 shows the proportion of patients whose concentration after the 36 h
dose reaches 20-60 mg/L, read at 38 h for regimens 1-3 and at 39 h for regimen
4 (the 1.5 g dose takes 2.5 h to infuse). As for Figure S1, the plotted points
were read exactly from the vector graphic. The paper concludes that none of the
regimens carries a risk of peaks above 40 mg/L.

```{r figS2}
figS2_pub <- tibble::tribble(
  ~regimen_short, ~time, ~conc, ~prop_pub,
  "R1", 38, 20, 0.992, "R1", 38, 30, 0.145, "R1", 38, 40, 0.000,
  "R2", 38, 20, 0.442, "R2", 38, 30, 0.000, "R2", 38, 40, 0.000,
  "R3", 38, 20, 0.928, "R3", 38, 30, 0.035, "R3", 38, 40, 0.000,
  "R4", 39, 20, 1.000, "R4", 39, 30, 0.677, "R4", 39, 40, 0.045
)
figS2_cmp <- sim |>
  dplyr::mutate(regimen_short = substr(regimen, 1, 2)) |>
  dplyr::inner_join(dplyr::distinct(figS2_pub, regimen_short, time),
                    by = c("regimen_short", "time")) |>
  tidyr::crossing(conc = c(20, 30, 40)) |>
  dplyr::group_by(regimen, regimen_short, time, conc) |>
  dplyr::summarise(prop_sim = mean(Cc >= conc), .groups = "drop") |>
  dplyr::inner_join(figS2_pub, by = c("regimen_short", "time", "conc"))

figS2_cmp |>
  dplyr::mutate(prop_sim = round(prop_sim, 2)) |>
  dplyr::select(regimen, time, conc, prop_pub, prop_sim) |>
  dplyr::rename(
    "Regimen" = regimen, "Time (h)" = time, "Threshold (mg/L)" = conc,
    "Published (Fig. S2)" = prop_pub, "Simulated" = prop_sim
  ) |>
  knitr::kable(caption = "Probability of a post-dose level at or above each threshold after the 36 h dose.")

stopifnot(
  median(abs(figS2_cmp$prop_sim - figS2_cmp$prop_pub)) < 0.08,
  # 'Not associated with a risk of toxic peak concentrations' (> 40 mg/L).
  all(figS2_cmp$prop_sim[figS2_cmp$conc == 40] < 0.15)
)
```

# Assumptions and deviations

- **Mean, not median, typical values.** Table 3 prints the mean, SD, CV% and
  median of each parameter's NPAG distribution. The means are encoded because
  the Results text quotes them and because they reproduce the paper's own PTA
  simulations (Figure 2) more closely than the medians (see above).
- **Log-normal approximation to a non-parametric distribution.** NPAG estimates
  a discrete joint distribution of support points, not a parametric omega. The
  CV% of Table 3 is carried as a log-normal variance, `omega^2 = log(CV^2 + 1)`.
  Any skew or multimodality of the true distribution (visible as the mean-median
  gap in V) cannot be recovered from the published table, and no CL-V
  correlation is reported, so the two random effects are independent.
- **Residual error not reported.** Neither the Pmetrics assay-error polynomial
  nor the fitted gamma/lambda term is given in the paper or in Additional file 1;
  both residual terms are carried as `fixed(0)`. Supply your own residual error
  before using the model to simulate observed TDM levels. The reproduction of
  Figures 2, S1 and S2 above without residual error indicates that the paper's
  PTA simulations were also of error-free concentrations.
- **Infusion rate in the PTA simulations.** The paper does not state the
  infusion duration used in its simulations; the ICU's 10 mg/min policy rate
  (Methods) is assumed. A 1-h infusion instead changes the simulated 38-39 h
  peaks by a few percentage points and the troughs and AUC24-48 negligibly.
- **Clearance is total clearance.** No effluent concentrations were measured, so
  CL lumps residual native and CVVHDF clearance; the model carries no CVVHDF
  settings (effluent, dialysate or blood flow) and applies only to patients
  receiving CVVHDF comparable to the study's (Prismaflex, PAES filter, blood
  flow 120-220 mL/min).
- **Covariate models not packaged.** Table 3 also reports models with clearance
  split by anticoagulation modality (regional citrate: CL 2.73 L/h; non-citrate:
  2.54 L/h) or by vasopressor use (with: 2.73 L/h; without: 2.53 L/h), and
  example continuous-covariate models (effluent flow, linear; flux,
  exponential; Additional file 1 Table S1 gives their equation forms). The
  authors did not select any of them, citing no practical difference from the
  base model, parameter correlations above 0.9 in the continuous-covariate
  models, and no gain in the other model-comparison metrics (Additional file 1
  Table S2); they used the base model for all dosing simulations. Only the
  base model is packaged.
- **Level counts.** The Abstract reports 155 levels and the Results text 74
  peaks plus 96 troughs (170); both are recorded as printed in the
  `population` metadata.
- **Digitisation.** The Figure 2 values were digitised by the maintainers from
  the published raster image (about +/-0.02); the Figure S1 and S2 values were
  read exactly from the vector drawing coordinates of Additional file 1.
- No erratum or correction to Kirwan 2021 was found as of 2026-09-30.
