---
title: "Cebranopadol (Kleideiter 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cebranopadol (Kleideiter 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Kleideiter E, Piana C, Wang S, Nemeth R, Gautrois M. Clinical pharmacokinetic characteristics of cebranopadol, a novel first-in-class analgesic. Clin Pharmacokinet. 2018;57(1):31-50.
* DOI: <https://doi.org/10.1007/s40262-017-0545-1>
* Article: <https://link.springer.com/article/10.1007/s40262-017-0545-1>
* Erratum: <https://doi.org/10.1007/s40262-018-0686-x> (Clin Pharmacokinet. 2018;57(11):1467-1469)

The erratum corrects Table 14 covariate-effect simulation results and several
paragraphs of the Section 3 discussion. Where erratum-corrected values differ
from the original article, the corrected values are the source of record for
the model and this vignette.

# Population

Kleideiter 2018 fits a two-compartment population PK model with two lagged
transition compartments and first-order elimination to 14 pooled clinical trials
of oral cebranopadol (a novel first-in-class NOP / opioid receptor agonist
analgesic) in healthy adults and adult chronic-pain patients:

* **Phase I (8 trials, n = 287 pooled).** Healthy adults receiving single oral
  doses of 0.8-800 ug or once-daily multiple doses up to 1600 ug. Three
  formulations (oral solution, film-coated tablet, liquid-filled capsule)
  were investigated and pooled in the population PK analysis (Table 12).
* **Phase IIa / II (6 trials, n = 1006 pooled).** Adult patients with chronic
  pain across four indications: chronic low back pain (cLBP, two trials),
  osteoarthritis (OA) of the knee, painful diabetic polyneuropathy (DPN, three
  trials), and post-bunionectomy acute pain (one trial), each at multiple dose
  levels with 4-14 weeks of once-daily dosing.

The pooled analysis cohort is 1293 subjects (617 men and 676 women per the
Discussion paragraph 4). Phase I subjects have median age 33 y (range 18-64)
and phase II subjects have median age 58 y (range 18-79); pooled weight range
is approximately 45-197 kg (Table 12 minimum to maximum across trials).
Genotype-derived CYP2C9 phenotype is available for 38.3% of subjects; the
remaining 61.7% are coded as `unknown` and form the model's CYP2C9 reference
category.

Reference covariate values (Kleideiter 2018 Table 14 footnote, erratum-corrected)
used for the parameter-estimation typical-value computation:

* sex = male (Table 14 simulation reference) / sex = female (Table 13
  parameter-estimation reference: the most-common category)
* formulation = tablet
* CYP2C9 status = unknown
* disease status = LBP and OA patients (nociceptive pain)
* age = 55 years
* CrCl = 106.4 mL/min
* body weight = 82 kg
* ALT = 19 U/L

Note that the Table 13 typical-value reference for sex (female) differs from
the Table 14 simulation reference (male). The model file encodes parameters
with female (SEXF = 1) as the reference and applies the +17.6% male-vs-female
CL ratio via `e_male_cl^(1 - SEXF)`; this matches the Table 13 typical-value
column directly and reproduces the Table 14 sex effect after the simulation
reference is switched.

Programmatic access to the population summary:
`readModelDb("Kleideiter_2018_cebranopadol")$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Kleideiter_2018_cebranopadol.R`. The table
below collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL/F, L/h) | log(74.3) | Kleideiter 2018 Table 13 Clearance Reference value |
| `lvc` (Vc/F, L) | log(225) | Kleideiter 2018 Table 13 Volume central compartment Reference value |
| `lvp` (Vp/F, L) | log(6750) | Kleideiter 2018 Table 13 Volume peripheral compartment Reference value |
| `lq` (Q/F, L/h) | log(84.2) | Kleideiter 2018 Table 13 Intercompartmental clearance |
| `lka` (Ka, 1/h) | log(0.864) | Kleideiter 2018 Table 13 Absorption rate constant Reference value |
| `lklag` (klag, 1/h) | log(0.087) | Kleideiter 2018 Table 13 klag Reference value |
| `e_male_cl` | 1.176 | Kleideiter 2018 Table 13 Male CL 87.4 / Female CL 74.3 |
| `e_em_cl` | 1.109 | Kleideiter 2018 Table 13 CYP2C9 EM CL 82.4 / Reference CL 74.3 |
| `e_pim_cl` | 0.790 | Kleideiter 2018 Table 13 CYP2C9 PIM CL 58.7 / Reference CL 74.3 |
| `e_alt_cl` | -0.156 | Kleideiter 2018 Table 13 Effect of ALT (exponential) |
| `e_crcl_cl` | 0.349 | Kleideiter 2018 Table 13 Effect of CrCl (exponential) |
| `e_age_vc` | -0.446 | Kleideiter 2018 Table 13 Effect of age (exponential) on Vc |
| `e_wt_vp` | 0.604 | Kleideiter 2018 Table 13 Effect of body weight (exponential) on Vp |
| `e_solution_ka` | 2.813 | Kleideiter 2018 Table 13 Oral-solution Ka 2.43 / Reference Ka 0.864 |
| `e_capsule_ka` | 2.419 | Kleideiter 2018 Table 13 Capsules Ka 2.09 / Reference Ka 0.864 |
| `e_solution_klag` | 0.885 | Kleideiter 2018 Table 13 Oral-solution klag 0.077 / Reference klag 0.087 |
| `e_capsule_klag` | 0.885 | Kleideiter 2018 Table 13 Capsules klag 0.077 / Reference klag 0.087 |
| `e_solution_f` | 1.045 | Kleideiter 2018 Table 13 Bioavailability Oral solution |
| `e_capsule_f` | 1.174 | Kleideiter 2018 Table 13 Bioavailability Capsules |
| `e_healthy_f` | 0.837 | Kleideiter 2018 Table 13 Bioavailability Healthy volunteers |
| `e_bun_f` | 1.801 | Kleideiter 2018 Table 13 (erratum-corrected; original Table-13 row labeled 'DPN patients') |
| `e_dpn_f` | 1.132 | Kleideiter 2018 Table 13 (erratum-corrected; original Table-13 row labeled 'Bunionectomy patients') |
| IIV CL (omega^2) | 0.412 (RSE 10.1%) | Kleideiter 2018 Table 13 IIV column |
| IIV Vc (omega^2) | 0.559 (RSE 20.6%) | Kleideiter 2018 Table 13 IIV column |
| IIV Ka (omega^2) | 0.519 (RSE 11.2%) | Kleideiter 2018 Table 13 IIV column |
| IIV klag (omega^2) | 0.0626 (RSE 18.8%) | Kleideiter 2018 Table 13 IIV column |
| `propSd` | 0.35 (approximated; see Assumptions) | Class-typical popPK value; Kleideiter 2018 does not report a residual-error magnitude in Table 13 or the Results narrative |
| Equations: 2-cmt + 2 transit + 1st-order elimination | n/a | Kleideiter 2018 Methods 2.4.1 + Results 3.2.7 |

# Virtual cohort

The original observed dataset is not publicly available. This vignette
constructs a typical-value simulation cohort matched to the Table 14 erratum
reference scenario (male, tablet, CYP2C9 unknown, LBP / OA, age 55 y,
CrCl 106.4 mL/min, weight 82 kg, ALT 19 U/L), plus the Table 14 covariate
perturbation arms (female sex, lower CrCl, healthy / DPN / bunionectomy disease
status), to reproduce the Table 14 erratum-corrected steady-state exposure
metrics under the published 600 ug titration regimen.

```{r virtual-pop}
set.seed(545L) # last 3 digits of original DOI 10.1007/s40262-017-0545-1

ref_covariates <- list(
  SEXF          = 0L,    # male = Table 14 simulation reference
  AGE           = 55,
  WT            = 82,
  CRCL          = 106.4,
  ALT           = 19,
  FORM_CAPSULE  = 0L,
  FORM_SOLUTION = 0L,
  CYP2C9_EM     = 0L,
  CYP2C9_PM_IM  = 0L,
  DIS_HEALTHY   = 0L,
  DIS_DPN       = 0L,
  DIS_BUN       = 0L
)

# One subject per arm, deterministic (zero between-subject variance) so the
# typical-value Cmax,ss and AUCss can be compared directly against Table 14
# erratum-corrected numbers.
make_arm <- function(arm_label, override = list(), id_offset = 0L) {
  cov <- modifyList(ref_covariates, override)
  data.frame(
    id            = id_offset + 1L,
    arm           = arm_label,
    SEXF          = cov$SEXF,
    AGE           = cov$AGE,
    WT            = cov$WT,
    CRCL          = cov$CRCL,
    ALT           = cov$ALT,
    FORM_CAPSULE  = cov$FORM_CAPSULE,
    FORM_SOLUTION = cov$FORM_SOLUTION,
    CYP2C9_EM     = cov$CYP2C9_EM,
    CYP2C9_PM_IM  = cov$CYP2C9_PM_IM,
    DIS_HEALTHY   = cov$DIS_HEALTHY,
    DIS_DPN       = cov$DIS_DPN,
    DIS_BUN       = cov$DIS_BUN
  )
}

arms <- bind_rows(
  make_arm("Reference (male, LBP/OA)",  id_offset =  0L),
  make_arm("Female",                    list(SEXF = 1L),                id_offset =  1L),
  make_arm("Age 40 y",                  list(AGE = 40),                 id_offset =  2L),
  make_arm("Age 75 y",                  list(AGE = 75),                 id_offset =  3L),
  make_arm("CrCl 45 mL/min",            list(CRCL = 45),                id_offset =  4L),
  make_arm("CrCl 60 mL/min",            list(CRCL = 60),                id_offset =  5L),
  make_arm("CrCl 80 mL/min",            list(CRCL = 80),                id_offset =  6L),
  make_arm("WT 70 kg",                  list(WT = 70),                  id_offset =  7L),
  make_arm("WT 100 kg",                 list(WT = 100),                 id_offset =  8L),
  make_arm("WT 120 kg",                 list(WT = 120),                 id_offset =  9L),
  make_arm("Healthy",                   list(DIS_HEALTHY = 1L),         id_offset = 10L),
  make_arm("DPN",                       list(DIS_DPN = 1L),             id_offset = 11L),
  make_arm("Bunionectomy",              list(DIS_BUN = 1L),             id_offset = 12L)
)
stopifnot(!anyDuplicated(arms$id))
n_arms <- nrow(arms)
```

# Simulation

Each arm receives the published 600 ug titration regimen (100 ug x 6 days,
200 ug x 6 days, 400 ug x 6 days, then 600 ug q24h) for 40 days of total
treatment, by which point the steady state is reached per the Table 14 erratum
footnote ("the steady-state was defined to be achieved on Day 40").

```{r build-events}
dose_schedule <- bind_rows(
  data.frame(day = 1:6,   amt = 100),
  data.frame(day = 7:12,  amt = 200),
  data.frame(day = 13:18, amt = 400),
  data.frame(day = 19:40, amt = 600)
)
dose_times <- (dose_schedule$day - 1) * 24

# Day 40 (last 24 h dosing interval) dense sampling for Cmax,ss and AUCss.
ss_dose_time   <- 39 * 24    # start of the day-40 dose
ss_obs_times   <- ss_dose_time + c(0, 0.5, 1, 2, 3, 4, 5, 6, 7, 8, 10, 12, 16, 20, 24)

# Trough sampling across the titration to show the approach to steady state.
trough_times   <- (seq(0, 40, by = 1)) * 24

all_obs_times  <- sort(unique(c(trough_times, ss_obs_times)))

# Replicate each arm row across dose events
dose_recs <- arms[rep(seq_len(n_arms), each = nrow(dose_schedule)), ] |>
  mutate(
    time = rep(dose_times,         times = n_arms),
    amt  = rep(dose_schedule$amt,  times = n_arms),
    evid = 1L,
    cmt  = "depot"
  )

obs_recs <- arms[rep(seq_len(n_arms), each = length(all_obs_times)), ] |>
  mutate(
    time = rep(all_obs_times, times = n_arms),
    amt  = 0,
    evid = 0L,
    cmt  = "central"
  )

events <- bind_rows(dose_recs, obs_recs) |>
  arrange(id, time, desc(evid)) |>
  select(id, time, amt, evid, cmt, arm,
         SEXF, AGE, WT, CRCL, ALT,
         FORM_CAPSULE, FORM_SOLUTION,
         CYP2C9_EM, CYP2C9_PM_IM,
         DIS_HEALTHY, DIS_DPN, DIS_BUN)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

```{r simulate}
mod <- readModelDb("Kleideiter_2018_cebranopadol")
# Typical-value (zero between-subject variance) simulation so the simulated
# Cmax,ss and AUCss can be compared head-to-head against the Table 14 erratum
# typical-patient predictions.
mod_typical <- rxode2::zeroRe(mod)
sim <- rxode2::rxSolve(mod_typical, events = events, keep = c("arm")) |>
  as.data.frame()
```

# Approach to steady state during the 600 ug titration

```{r titration-vpc}
trough_sim <- sim |>
  filter(time %in% trough_times) |>
  mutate(day = time / 24)

ggplot(
  trough_sim |> filter(arm == "Reference (male, LBP/OA)"),
  aes(day, Cc)
) +
  geom_line(linewidth = 0.8, colour = "steelblue") +
  geom_vline(xintercept = c(6, 12, 18), linetype = "dashed", colour = "grey60") +
  annotate("text", x = c(3, 9, 15, 30), y = max(trough_sim$Cc) * 0.9,
           label = c("100 ug", "200 ug", "400 ug", "600 ug"), size = 3) +
  labs(
    x       = "Day after first dose",
    y       = "Trough cebranopadol concentration (pg/mL)",
    title   = "Typical-value trough profile during the 600 ug titration",
    caption = "Reference covariate set: male, tablet, CYP2C9 unknown, LBP/OA, age 55 y, CrCl 106.4 mL/min, WT 82 kg, ALT 19 U/L."
  ) +
  theme_bw()
```

# Replicate Kleideiter 2018 Table 14 (erratum-corrected) covariate effects

Table 14 (erratum) tabulates typical-patient Cmax,ss and AUCss values under
the 600 ug titration regimen for each covariate-perturbation arm, expressed
as % change relative to the reference patient.

```{r table14-nca}
# Day 40 (last 24-h interval) dense observations, time shifted so each arm has
# a t = 0 at the start of the day-40 dose.
ss_block <- sim |>
  filter(time %in% ss_obs_times) |>
  mutate(time = time - ss_dose_time) |>
  select(id, time, Cc, arm)

# Dose record at t = 0 of the shifted interval (600 ug on the steady-state day).
dose_block <- ss_block |>
  group_by(id) |>
  summarise(time = 0, amt = 600, arm = first(arm), .groups = "drop")

conc_obj <- PKNCA::PKNCAconc(
  ss_block,
  Cc ~ time | arm + id,
  concu = "pg/mL", timeu = "hr"
)
dose_obj <- PKNCA::PKNCAdose(
  dose_block,
  amt ~ time | arm + id,
  doseu = "ug"
)

intervals <- data.frame(
  start    = 0,
  end      = 24,
  cmax     = TRUE,
  auclast  = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

table14_sim <- as.data.frame(nca_res$result) |>
  filter(PPTESTCD %in% c("cmax", "auclast")) |>
  select(arm, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  rename(Cmax_ss_sim = cmax, AUC_ss_sim = auclast)

ref_row <- table14_sim |> filter(arm == "Reference (male, LBP/OA)")
ref_cmax <- ref_row$Cmax_ss_sim
ref_auc  <- ref_row$AUC_ss_sim

table14_sim <- table14_sim |>
  mutate(
    Cmax_pct_sim = 100 * (Cmax_ss_sim - ref_cmax) / ref_cmax,
    AUC_pct_sim  = 100 * (AUC_ss_sim  - ref_auc ) / ref_auc
  )

# Published Table 14 erratum values (Cmax,ss in pg/mL, AUCss in pg*h/mL, percent
# change relative to the male reference patient).
table14_pub <- tibble::tribble(
  ~arm,                       ~Cmax_ss_pub, ~Cmax_pct_pub, ~AUC_ss_pub, ~AUC_pct_pub,
  "Reference (male, LBP/OA)",        378,          0,           6790,         0,
  "Female",                          432,         14,           7960,        17,
  "Age 40 y",                        375,         -1,           6790,         0,
  "Age 75 y",                        381,          1,           6790,         0,
  "CrCl 45 mL/min",                  483,         28,           9080,        34,
  "CrCl 60 mL/min",                  446,         18,           8250,        21,
  "CrCl 80 mL/min",                  410,          9,           7480,        10,
  "WT 70 kg",                        379,          0,           6810,         0,
  "WT 100 kg",                       377,          0,           6760,        -1,
  "WT 120 kg",                       375,         -1,           6720,        -1,
  "Healthy",                         316,        -16,           5690,       -16,
  "DPN",                             428,         13,           7690,        13,
  "Bunionectomy",                    681,         80,          12200,        80
)

table14_compare <- table14_pub |>
  left_join(table14_sim, by = "arm") |>
  select(arm,
         Cmax_ss_pub, Cmax_ss_sim, Cmax_pct_pub, Cmax_pct_sim,
         AUC_ss_pub,  AUC_ss_sim,  AUC_pct_pub,  AUC_pct_sim)

knitr::kable(
  table14_compare,
  digits  = c(0, 0, 0, 0, 1, 0, 0, 0, 1),
  caption = paste(
    "Simulated typical-value Cmax,ss (pg/mL) and AUCss (pg*h/mL) under the",
    "600 ug titration regimen vs Kleideiter 2018 Table 14 erratum-corrected",
    "published values. Percent changes are relative to the reference patient."
  )
)
```

The simulated percent-change column should reproduce the Table 14 erratum
percent-change column closely for the categorical covariate arms (sex,
disease status); the absolute Cmax,ss and AUCss values may differ from the
published numbers because the structural interpretation of the transit-
compartment absorption chain affects the entire concentration-time profile
shape. See Assumptions and deviations for details.

# Assumptions and deviations

- **Transit-compartment structural interpretation.** Kleideiter 2018 Section
  3.2.7 describes the absorption model as "a two-compartment disposition model
  with two lagged transition compartments" without writing out the explicit
  ODEs, and Table 13 reports two formulation-dependent rate constants `Ka` and
  `klag` whose individual roles in the chain are not stated. The model file
  uses the biologically-motivated reading in which `Ka` governs the depot-to-
  transit1 step (dissolution-limited absorption from the dosage form; values
  vary ~2.8-fold across tablet, capsule, and oral solution) and `klag` governs
  both subsequent transit steps (transit1 -> transit2 -> central; uniform 0.077-
  0.087 1/h across formulations). The alternative reading - `klag` at the
  depot + transit1 steps and `Ka` at the final transit2 -> central step (the
  Bienczak 2016 nevirapine convention used elsewhere in nlmixr2lib) - is
  equally consistent with the paper's prose; without the NONMEM control
  stream or appendix, the ambiguity cannot be resolved from on-disk sources.
  Either reading places the rate-limiting step at the slow `klag` rate, so
  the simulated peak time and concentration-time-profile shape will not
  precisely reproduce the paper's reported Tmax of 4-6 h (the NCA-derived
  Tmax from the single-dose trials). The Table 14 covariate-percent-change
  comparison is more robust to this structural ambiguity than the absolute
  Cmax,ss / AUCss numbers because percent changes within an arm cancel the
  common transit-time scale.
- **Residual-error magnitude.** Kleideiter 2018 Section 2.4.1 specifies the
  residual-error model as additive on log-transformed cebranopadol
  concentrations (equivalent to proportional residual error on the linear
  scale in the nlmixr2 idiom), but Table 13 and the Results narrative do
  not report the residual-error variance estimate. The model file uses a
  class-typical approximation `propSd = 0.35` (about 35% proportional CV
  on the linear scale, a representative value for orally-dosed
  small-molecule analgesics with mixed intensive + sparse-sampling popPK
  datasets). This value affects the width of stochastic VPCs (irrelevant
  for the deterministic Table 14 reproduction above) but not the typical-
  value Cmax,ss / AUCss predictions.
- **Sex reference category divergence between Table 13 and Table 14.** The
  Table 13 parameter-estimation reference uses female sex (the most-common
  category in the analysis cohort), while the Table 14 simulation reference
  uses male sex (the typical-patient simulation choice). The model file
  preserves the Table 13 reference - female corresponds to SEXF = 1, the
  typical-value CL is 74.3 L/h, and male is encoded as the +17.6%
  multiplier applied via `e_male_cl^(1 - SEXF)`. This vignette's
  reference-arm simulation switches to the Table 14 male reference for
  side-by-side comparison.
- **CYP2C9 unknown-phenotype reference.** Only 38.3% of the analysis cohort
  had a known CYP2C9 phenotype; the remaining 61.7% are coded as `unknown`
  and form the model's CYP2C9 reference category. The canonical
  `CYP2C9_EM` covariate's standard reference (in Jeong 2022) is IM/PM only,
  not unknown. In the Kleideiter 2018 model, the 0-level of `CYP2C9_EM`
  pools unknown + non-EM phenotypes, and the sibling canonical
  `CYP2C9_PM_IM` carries the PM/IM-vs-unknown contrast. Both indicators are 0
  for the reference-arm simulations above.
- **Four-level disease-state encoding via three indicators.** The model
  encodes the four-category disease factor (LBP and OA reference + healthy,
  DPN, bunionectomy) using three binary indicators `DIS_HEALTHY`,
  `DIS_DPN`, and `DIS_BUN`, two of which (`DIS_DPN`, `DIS_BUN`) are
  registered as new canonical covariates alongside this model. When all
  three are 0 the subject is in the LBP / OA reference category (the
  Table 13 most-common-category reference).
- **Three-level formulation encoding via two indicators.** The model
  encodes the three-category formulation factor (tablet reference + oral
  solution + liquid-filled capsule) using two binary indicators
  `FORM_SOLUTION` (new canonical registered alongside this model) and
  `FORM_CAPSULE` (existing canonical, reused with the tablet reference
  documented per-model). When both are 0 the subject is in the tablet
  reference category.
- **Food and CYP2D6 effects not retained.** Kleideiter 2018 Methods
  Section 2.4.1 tested food intake on the absorption rate constant and
  CYP2D6 phenotype on CL during covariate selection; neither was retained
  in the final model (Results 3.2.7 paragraphs 5-7). The model file
  therefore does not include `FED` or `CYP2D6_*` covariates. Future
  cebranopadol simulations that require a food-effect term should consult
  the paper's Section 3.2.6 single-dose ANOVA estimates (mean Cmax and
  AUCt ~30% higher under fed vs fasted conditions) rather than the
  population PK model.
- **Erratum supersedes original Table 14.** The Cmax,ss / AUCss percent-
  change values in the published-vs-simulated comparison above are taken
  from the 2018 erratum's revised Table 14, not the original 2018 article
  Table 14, per the Phase 1 step 8 errata-handling convention.
