---
title: "Alprazolam (Klein 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alprazolam (Klein 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Klein_2026_alprazolam"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1111/epi.18643>
* Supporting Information (Tables S1-S3, Figures S1-S5): distributed with the
  article; retrieved from the Europe PMC supplementary-file archive for
  PMC12893244.

Staccato alprazolam is a hand-held, breath-actuated thermal-aerosol device that
vaporises alprazolam and delivers it to the deep lung, where it is absorbed at
the alveolar-capillary interface. The population PK model reported here pools
adolescent data from the Phase 1 trial UP0100 with adult data from two earlier
trials, and was used to support the 2 mg adolescent dose in the Phase 3
programme.

# Population

```{r population, echo = FALSE}
pop <- ui$population
tibble::tibble(Field = names(pop),
               Value = vapply(pop, \(x) paste(
                 if (is.null(names(x))) x else paste0(names(x), ": ", x),
                 collapse = "; "), character(1))) |>
  knitr::kable(caption = "Population metadata recorded with the model.")
```

The analysis dataset (Table S1) pools `r pop$n_subjects` subjects from
`r pop$n_studies` trials: UP0100 (Phase 1, 14 adolescents aged 12-17 years with
focal, generalized, or focal and generalized epilepsy, NCT04857307),
AMDC-002-202 (Phase 2a, 5 adults with photosensitive epilepsy, NCT02351115) and
ENGAGE-E-001 (Phase 2b, 80 adults with epilepsy, NCT03478982). Overall mean age
was 32.2 years (SD 13.5) and mean weight 75.5 kg (SD 22.5); 70.7% were female.
Fifteen subjects (15.2%), all in ENGAGE-E-001, were taking a strong hepatic
enzyme-inducing antiseizure medication; no adolescent was, because UP0100
excluded patients on strong CYP3A4 inducers by protocol.

# Model structure

Absorption is modelled as **two parallel processes whose fractions sum to one**:

* an **immediate-release** fraction that appears in `central` as a bolus at the
  dose time (the paper's `F2` = 0.369, the "fast absorption fraction"), and
* the complementary fraction (0.631) that enters `depot` and is absorbed
  first-order with `ka` = 5.37 /h, i.e. an absorption half-life of
  `r sprintf("%.2f", log(2) / 5.37 * 60)` min.

Because both arms are fed from the *same* administered dose, an event table for
this model must carry **two dose records at each administration time** -- one to
`depot` and one to `central`. The `f(depot)` / `f(central)` terms then apportion
them. Every cohort below follows that pattern.

Disposition is two-compartment with linear elimination. Body weight scales the
central and peripheral volumes through a single shared exponent; it does **not**
scale CL or Q, because the authors could detect no weight relationship for
either (Methods section 2.2, and the title of Table S3).

# Source trace

Every value below is also carried as an in-file comment beside its `ini()` entry
in `inst/modeldb/specificDrugs/Klein_2026_alprazolam.R`.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lka` | 5.37 /h (95% CI 3.88-7.45) | Table S3, row `ka`; absorption half-life 7.74 (5.58-10.7) min restated in Results section 3.2 |
| `lcl` | 8.23 L/h (7.23-9.38) | Table S3, row `CL/F`; restated in Results section 3.2 |
| `lvc` | 36.9 L (23.1-59.2) | Table S3, row `Vc/F` |
| `lq` | 189 L/h (131-271) | Table S3, row `Q/F` |
| `lvp` | 60.3 L (47.2-77.0) | Table S3, row `Vp/F` |
| `logitfdepot` | `qlogis(1 - 0.369)` | Table S3, row `F2` = 0.369 (0.156-0.649); restated in Results section 3.2 |
| `e_wt_vc_vp` | 0.470 (0.341-0.598) | Table S3, row `Allometric scaling Vc and Vp` |
| `e_conmed_eiaed_cl` | `log(1.955)` | Table S3, row `Effect of inducer ASMs on CL, % change` = 95.5 (50.5-154.0) |
| `etalka` | 0.730^2 | Table S3, IIV column, row `ka` = 73.0% |
| `etalcl`, `etalvc` block | 0.586^2, 0.763^2, r = 0.342 | Table S3, IIV column rows `CL/F` 58.6% and `Vc/F` 76.3%; footnote "Correlation between CL and Vc: 0.342" |
| `etalq` | 0.396^2 | Table S3, IIV column, row `Q/F` = 39.6% |
| `etalvp` | 0.209^2 | Table S3, IIV column, row `Vp/F` = 20.9% |
| `etalogitfdepot` | 1.078^2 | Table S3, IIV column, row `F2` = 107.8% |
| `propSd` | 0.100 | Table S3, row `Proportional RUV, %` = 10.0 (8.30-11.7) |
| Two parallel absorption arms; 2-compartment disposition; linear elimination | n/a | Methods section 2.2, paragraph 3 |
| No weight effect on CL or Q; shared allometric exponent on Vc and Vp | n/a | Methods section 2.2, paragraph 4, and Table S3 title |
| `WT` reference 70 kg | assumed | **not stated in the paper**; see Assumptions below |

## How each estimation scale was identified

Table S3 reports a 95% CI for every structural parameter but never states the
scale each was estimated on. Each CI is Wald-symmetric on exactly one scale, so
the CIs identify the scales without any appeal to convention. Recomputing the
midpoint on each candidate scale and comparing with the published point estimate:

```{r scale-check}
ci <- tibble::tribble(
  ~parameter,  ~estimate, ~lower,  ~upper,
  "ka",             5.37,   3.88,    7.45,
  "CL/F",           8.23,   7.23,    9.38,
  "Vc/F",          36.9,   23.1,    59.2,
  "Q/F",          189,    131,     271,
  "Vp/F",          60.3,   47.2,    77.0,
  "F2",             0.369,  0.156,   0.649,
  "Allom. exp.",    0.470,  0.341,   0.598,
  "Prop. RUV (%)", 10.0,    8.30,   11.7
)

# The logit scale is only defined for a bound in (0, 1); return NA elsewhere
# rather than letting qlogis() emit NaN.
logit_mid <- function(lower, upper) {
  ok <- lower > 0 & lower < 1 & upper > 0 & upper < 1
  ifelse(ok, plogis((qlogis(ifelse(ok, lower, 0.5)) +
                       qlogis(ifelse(ok, upper, 0.5))) / 2), NA_real_)
}

scale_check <- ci |>
  mutate(
    linear = (lower + upper) / 2,
    log    = exp((log(lower) + log(upper)) / 2),
    logit  = logit_mid(lower, upper),
    err    = pmin(abs(linear / estimate - 1),
                  abs(log / estimate - 1),
                  abs(logit / estimate - 1), na.rm = TRUE),
    best   = c("linear", "log", "logit")[
      apply(cbind(abs(linear / estimate - 1),
                  abs(log / estimate - 1),
                  abs(logit / estimate - 1)), 1,
            \(x) which.min(replace(x, is.na(x), Inf)))]
  )

scale_check |>
  mutate(across(c(linear, log, logit), \(x) signif(x, 4))) |>
  dplyr::rename("Parameter" = parameter, "Published estimate" = estimate,
                "Midpoint, linear" = linear, "Midpoint, log" = log,
                "Midpoint, logit" = logit, "Scale identified" = best) |>
  dplyr::select(-lower, -upper, -err) |>
  knitr::kable(caption = "CI midpoint on each candidate scale vs the published point estimate. The scale whose midpoint reproduces the estimate is the scale the parameter was estimated on.")
```

`F2` is only interpretable on the **logit** scale: its logit midpoint reproduces
0.369 to three decimals, while the linear midpoint gives 0.4025 and the log
midpoint 0.3182. That is decisive, and it is corroborated structurally -- a
log-normal IIV of the reported magnitude on a fraction bounded by 1 would push
roughly 13% of subjects above `F2` = 1, i.e. a negative depot fraction. The
model therefore carries the dose fraction on the logit scale, using the
library's `logitfdepot` idiom.

The inducer effect is likewise a log-scale coefficient: the midpoint of
`log(1.505)` and `log(2.540)` is `r sprintf("%.5f", (log(1.505) + log(2.540)) / 2)`,
which is `log(1.955)` to five decimals, whereas the linear midpoint of the
reported percentages is `r sprintf("%.2f", (50.5 + 154.0) / 2)`%, not 95.5%.

```{r scale-assert}
# Every parameter's estimation scale must be recovered from its own CI.
stopifnot(identical(
  scale_check$best,
  c("log", "log", "log", "log", "log", "logit", "linear", "linear")
))
# ...and on that scale the midpoint must reproduce the estimate to within 0.5%.
stopifnot(!anyNA(scale_check$err), all(scale_check$err < 0.005))
```

## Which reading of the "IIV" column?

Table S3 reports interindividual variability as a bare percentage with no
back-transformation stated. Two readings are current in the literature and they
differ by about 8% on omega:

* omega is the **standard deviation** on the parameter's own estimation scale,
  reported as `omega * 100` -- for CL that gives omega = 0.586;
* the percentage is the **exact log-normal CV%**, `sqrt(exp(omega^2) - 1) * 100`
  -- for CL that gives omega = 0.543.

Figure 4 settles it in closed form, with no simulation and no digitising of a
curve. The Supplementary Methods state that AUCinf was computed *"using the
sampled clearance (CL) parameter by dividing the administered dose by CL"*, so
the simulated AUCinf distribution is exactly log-normal with median `dose / CL`
and 5th/95th percentiles `median * exp(-/+ 1.645 * omega_CL)`. Only one reading
can match the published band.

```{r iiv-scale}
dose_ug   <- 2000          # 2 mg expressed in ug so AUC comes out in ng*h/mL
cl_typ    <- 8.23
inducer   <- 1.955         # CL multiplier with an enzyme-inducing ASM

omega_sd  <- 0.586                          # "IIV % is omega * 100"
omega_cv  <- sqrt(log(1 + 0.586^2))         # "IIV % is the exact log-normal CV%"

band <- function(omega, cl) {
  med <- dose_ug / cl
  c(p05 = med / exp(1.645 * omega), median = med, p95 = med * exp(1.645 * omega))
}

# Values read off Figure 4 (blue band and red median line, adolescent panels).
digitised <- tibble::tribble(
  ~panel,          ~p05, ~median, ~p95,
  "No inducer ASM",  93,     240,  645,
  "Inducer ASM",     45,     122,  330
)

iiv_cmp <- bind_rows(
  tibble::tibble(panel = "No inducer ASM", reading = "omega = SD (0.586)",
                 !!!band(omega_sd, cl_typ)),
  tibble::tibble(panel = "No inducer ASM", reading = "omega = exact CV% (0.543)",
                 !!!band(omega_cv, cl_typ)),
  tibble::tibble(panel = "Inducer ASM", reading = "omega = SD (0.586)",
                 !!!band(omega_sd, cl_typ * inducer)),
  tibble::tibble(panel = "Inducer ASM", reading = "omega = exact CV% (0.543)",
                 !!!band(omega_cv, cl_typ * inducer))
) |>
  left_join(digitised, by = "panel", suffix = c("", "_fig")) |>
  mutate(`error, 5th (%)`  = 100 * (p05 / p05_fig - 1),
         `error, 95th (%)` = 100 * (p95 / p95_fig - 1))

iiv_cmp |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  dplyr::select(panel, reading, p05, p95, p05_fig, p95_fig,
                `error, 5th (%)`, `error, 95th (%)`) |>
  dplyr::rename("Figure 4 panel" = panel, "Reading of the IIV column" = reading,
                "5th, predicted" = p05, "95th, predicted" = p95,
                "5th, Figure 4" = p05_fig, "95th, Figure 4" = p95_fig) |>
  knitr::kable(caption = "AUCinf (ng*h/mL) 90% range under each reading of Table S3's IIV column, against values read off Figure 4.")
```

```{r iiv-scale-assert}
err <- iiv_cmp |>
  group_by(reading) |>
  summarise(mean_abs_err = mean(abs(c(`error, 5th (%)`, `error, 95th (%)`))),
            .groups = "drop")
# The SD reading must win, and by a clear margin, on the pooled tail error.
stopifnot(
  err$mean_abs_err[err$reading == "omega = SD (0.586)"] <
    err$mean_abs_err[err$reading == "omega = exact CV% (0.543)"],
  err$mean_abs_err[err$reading == "omega = SD (0.586)"] < 5
)
# The median is common to both readings and must land on the published line.
stopifnot(abs(dose_ug / cl_typ / 240 - 1) < 0.05,
          abs(dose_ug / (cl_typ * inducer) / 122 - 1) < 0.05)
```

The SD reading reproduces both tails of both panels to within a few percent; the
exact-CV% reading is systematically too narrow, missing the no-inducer 95th
percentile by about 8%. The model therefore encodes `omega = IIV% / 100` on each
parameter's own estimation scale. The two readings differ by less than the width
of the published CIs for the structural parameters, so this choice matters for
simulated variability, not for typical-value predictions.

# Structural identity checks

These four checks are deterministic: they compare the solved model against
closed forms implied by the published parameters, with the random effects zeroed.
They contain no cohort sampling, so they can be asserted tightly.

```{r identities}
mod     <- readModelDb("Klein_2026_alprazolam")
mod_typ <- rxode2::zeroRe(mod)

# Two dose records per administration: the immediate-release arm and the depot arm.
dose_rows <- function(id, wt, eiaed) {
  tibble::tibble(id = id, WT = wt, CONMED_EIAED = eiaed, time = 0,
                 amt = 2, evid = 1L, cmt = c("depot", "central"))
}
obs_rows <- function(id, wt, eiaed, times) {
  tibble::tibble(id = id, WT = wt, CONMED_EIAED = eiaed, time = times,
                 amt = NA_real_, evid = 0L, cmt = "central")
}

wt_bands <- c(`<50 kg` = 41.27, `>=50 kg` = 64.35, `All patients` = 54.46)
grid     <- seq(0, 36, length.out = 20001)

ev_typ <- bind_rows(lapply(seq_along(wt_bands), function(i) {
  bind_rows(dose_rows(i, wt_bands[[i]], 0), obs_rows(i, wt_bands[[i]], 0, grid))
})) |> arrange(id, time, desc(evid))

sim_typ <- rxode2::rxSolve(mod_typ, ev_typ, omega = NA, sigma = NA,
                           keep = c("WT")) |> as.data.frame()
stopifnot(nrow(sim_typ) > 0, all(sim_typ$Cc >= 0))

terminal_slope <- function(d) {
  -coef(stats::lm(log(Cc) ~ time, data = d[d$time >= 12, ]))[[2]]
}
auc_inf <- function(d) {
  a <- sum(diff(d$time) * (head(d$Cc, -1) + tail(d$Cc, -1)) / 2)
  a + tail(d$Cc, 1) / terminal_slope(d)
}

published_t12 <- c(`<50 kg` = 6.816, `>=50 kg` = 8.477, `All patients` = 7.601)

ident <- lapply(seq_along(wt_bands), function(i) {
  d  <- sim_typ[sim_typ$id == i, ]
  vc <- 36.9 * (wt_bands[[i]] / 70)^0.470
  tibble::tibble(
    band            = names(wt_bands)[i],
    WT              = wt_bands[[i]],
    t12_model       = log(2) / terminal_slope(d),
    t12_published   = published_t12[[i]],
    c0_solved       = d$Cc[1],
    c0_closed_form  = 0.369 * 2 / vc * 1000,
    aucinf_solved   = auc_inf(d),
    aucinf_dose_cl  = dose_ug / cl_typ
  )
}) |> bind_rows() |>
  mutate(`t1/2 error (%)` = 100 * (t12_model / t12_published - 1),
         `C(0+) ratio`    = c0_solved / c0_closed_form,
         `AUCinf ratio`   = aucinf_solved / aucinf_dose_cl)

ident |>
  mutate(across(where(is.numeric), \(x) signif(x, 6))) |>
  dplyr::select(band, WT, t12_model, t12_published, `t1/2 error (%)`,
                `C(0+) ratio`, `AUCinf ratio`) |>
  dplyr::rename("Weight band" = band, "WT (kg)" = WT,
                "t1/2, model (h)" = t12_model,
                "t1/2, Table 1 (h)" = t12_published) |>
  knitr::kable(caption = "Deterministic structural checks against closed forms and against the observed terminal half-lives in Table 1.")
```

```{r identities-assert}
# 1. The immediate-release bolus must be exactly F2 * dose / Vc(WT).
stopifnot(max(abs(ident$`C(0+) ratio` - 1)) < 1e-6)

# 2. AUCinf must be exactly dose / CL, and independent of weight, because the
#    paper found no weight effect on CL.
#    (relative tolerances: the comparator is a trapezoidal integral, so the
#    residual is numerical-quadrature error, not model error)
stopifnot(max(abs(ident$`AUCinf ratio` - 1)) < 1e-4,
          diff(range(ident$aucinf_solved)) / mean(ident$aucinf_solved) < 1e-4)

# 3. The model's terminal half-lives must reproduce the observed ones in Table 1,
#    including the ORDERING across weight bands -- that ordering is generated
#    purely by the allometric exponent on Vc and Vp, since CL is not scaled.
stopifnot(max(abs(ident$`t1/2 error (%)`)) < 10)
stopifnot(ident$t12_model[ident$band == "<50 kg"] <
            ident$t12_model[ident$band == ">=50 kg"])

# 4. The enzyme-inducer effect must divide exposure by exactly 1.955.
ev_ind <- bind_rows(dose_rows(1L, 70, 1), obs_rows(1L, 70, 1, grid)) |>
  arrange(time, desc(evid))
ev_non <- bind_rows(dose_rows(1L, 70, 0), obs_rows(1L, 70, 0, grid)) |>
  arrange(time, desc(evid))
auc_ind <- auc_inf(as.data.frame(rxode2::rxSolve(mod_typ, ev_ind, omega = NA, sigma = NA)))
auc_non <- auc_inf(as.data.frame(rxode2::rxSolve(mod_typ, ev_non, omega = NA, sigma = NA)))
stopifnot(abs((auc_non / auc_ind) / 1.955 - 1) < 1e-4)
```

The half-life check is the strongest single piece of evidence for the assumed
70 kg reference weight. `CL` is not weight-scaled, so the *only* thing that can
make terminal half-life vary with body weight in this model is the shared
allometric exponent on `Vc` and `Vp`. The model reproduces both the direction and
the magnitude of the observed difference between the two adolescent weight bands
(Table 1: 6.816 h below 50 kg vs 8.477 h at or above 50 kg) without any parameter
having been fitted to those numbers.

# Virtual adolescent cohort

The cohort reproduces the UP0100 demographics (Table S2): 200 adolescents with
body weight drawn from a normal distribution with the observed mean 54.46 kg and
SD 14.43 kg, truncated to the observed range 33.5-81.2 kg. None receives an
enzyme-inducing antiseizure medication, matching the trial's exclusion criterion.
Samples are taken at the **protocol times** -- pre-dose and 2, 10 and 30 min and
1, 2, 6, 24 and 36 h post-dose -- because Cmax and Tmax in Table 1 are discrete
statistics limited by that schedule, a limitation the authors call out explicitly
in the Discussion.

```{r cohort}
set.seed(20260828)
n_adol <- 200

wt <- rnorm(n_adol, 54.46, 14.43)
wt <- pmin(pmax(wt, 33.5), 81.2)

subj <- tibble::tibble(
  id           = seq_len(n_adol),
  WT           = wt,
  wtgrp        = ifelse(wt < 50, "<50 kg", ">=50 kg"),
  CONMED_EIAED = 0
)

protocol_times <- c(0, 2 / 60, 10 / 60, 0.5, 1, 2, 6, 24, 36)

events <- bind_rows(
  subj |> mutate(time = 0, amt = 2, evid = 1L, cmt = "depot"),
  subj |> mutate(time = 0, amt = 2, evid = 1L, cmt = "central"),
  tidyr::crossing(subj, time = protocol_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
) |>
  arrange(id, time, desc(evid))

# The two dose records at t = 0 must be distinguished by their target
# compartment, and the t = 0 observation by its evid. Check the events
# themselves -- unique()-ing first would make this assertion unfalsifiable.
stopifnot(
  anyDuplicated(events[, c("id", "time", "evid", "cmt")]) == 0L,
  # every subject carries exactly two dose records: one per absorption arm
  nrow(dplyr::filter(events, evid == 1L)) == 2L * n_adol
)
```

```{r simulate}
sim <- rxode2::rxSolve(mod, events, keep = c("WT", "wtgrp")) |> as.data.frame()
stopifnot(nrow(sim) > 0, dplyr::n_distinct(sim$id) == n_adol)
```

`sim` carries the observation with residual error and `Cc` the individual
prediction; the NCA below uses `sim`, because the published noncompartmental
analysis was run on measured concentrations, which carry the same 10%
proportional error.

# Replicate Figure 1

```{r figure-1, fig.cap = "Replicates Figure 1 of Klein 2026: geometric mean plasma alprazolam concentration-time profiles by adolescent body weight group after a single 2 mg dose."}
LLOQ <- 0.200

gm <- function(x) {
  x <- x[is.finite(x) & x > 0]
  if (!length(x)) return(NA_real_)
  exp(mean(log(x)))
}

sim |>
  filter(time > 0) |>
  group_by(wtgrp, time) |>
  summarise(geomean = gm(sim), .groups = "drop") |>
  ggplot(aes(time, geomean, colour = wtgrp, shape = wtgrp)) +
  geom_line() + geom_point() +
  geom_hline(yintercept = LLOQ, linetype = "dashed", linewidth = 0.3) +
  scale_y_log10() +
  scale_x_continuous(breaks = seq(0, 36, by = 6)) +
  labs(x = "Time (hours)", y = "Geometric mean concentration (ng/mL)",
       colour = "Body weight", shape = "Body weight",
       title = "Figure 1 (semilogarithmic) -- simulated adolescent cohort",
       caption = "Replicates Figure 1B of Klein 2026. Dashed line: LLOQ 0.200 ng/mL.")
```

As in the published figure, the lower weight band runs above the higher one over
the absorption and early distribution phases and crosses below it during the
terminal phase, the latter because only the volumes -- not clearance -- scale
with body weight.

# PKNCA validation

Concentrations below the 0.200 ng/mL lower limit of quantification are censored,
so `auclast` is computed to the last *quantifiable* concentration exactly as the
published `AUC0-t` was.

The immediate-release arm puts drug in `central` at the instant of dosing, so the
model's `time = 0` observation is a *post*-bolus concentration. The trial's
pre-dose sample was taken before administration, in drug-naive patients, and was
therefore zero. The pre-dose record is set to zero here to match, which also
keeps `Cmax` and `Tmax` derived from post-dose samples only, exactly as in the
trial.

```{r pknca}
post_dose <- sim |>
  dplyr::filter(time > 0) |>
  dplyr::select(id, time, wtgrp, Cc = sim) |>
  mutate(Cc = ifelse(Cc < LLOQ, NA_real_, Cc)) |>
  dplyr::filter(!is.na(Cc))

pre_dose <- subj |> transmute(id, wtgrp, time = 0, Cc = 0)

conc_df <- bind_rows(pre_dose, post_dose) |> arrange(id, time)

stopifnot(
  nrow(conc_df) > 0,
  # every subject must carry exactly one time-zero anchor record
  nrow(dplyr::filter(conc_df, time == 0)) == n_adol
)

dose_df <- subj |> transmute(id, wtgrp, time = 0, amt = 2)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(conc_df), Cc ~ time | wtgrp + id)
dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | wtgrp + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE,
  half.life = TRUE, cl.obs = TRUE, vss.obs = TRUE
)

nca_res <- suppressWarnings(
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
)
nca_df <- as.data.frame(nca_res)
stopifnot(nrow(nca_df) > 0)
```

`PKNCA` reports `cl.obs` and `vss.obs` in dose units per concentration unit --
here mg per (ng/mL), i.e. 1000 L -- so both are rescaled to litres below before
comparison with the published `CL/F` and `Vss/F`.

```{r nca-rescale}
nca_scaled <- nca_df |>
  mutate(PPORRES = ifelse(PPTESTCD %in% c("cl.obs", "vss.obs"),
                          PPORRES * 1000, PPORRES),
         PPORRES = ifelse(PPTESTCD == "tmax", PPORRES * 60, PPORRES))
```

## Comparison against published NCA

Table 1 of the paper reports **geometric means** for every parameter except
`Tmax`, which is a median. `ncaComparisonTable()` aggregates a per-subject input
with `median()`, so the simulated results are pre-aggregated here with the
matching statistic -- geometric mean throughout, median for `Tmax` -- and the
already-summarised frame is passed in. Comparing a simulated median against a
published geometric mean would score the two sides on different quantities.

```{r nca-compare}
published <- tibble::tribble(
  ~wtgrp,     ~cmax, ~aucinf.obs, ~auclast, ~cl.obs, ~tmax, ~vss.obs, ~half.life,
  "<50 kg",   38.27,       281.1,    270.8,   7.114,  19.50,    67.53,      6.816,
  ">=50 kg",  33.55,       278.8,    283.8,   7.174,  10.00,    87.38,      8.477
)

nca_summary <- nca_scaled |>
  group_by(wtgrp, PPTESTCD) |>
  summarise(PPORRES = if (first(PPTESTCD) == "tmax") {
              median(PPORRES, na.rm = TRUE)
            } else {
              gm(PPORRES)
            }, .groups = "drop")
stopifnot(nrow(nca_summary) > 0, !anyNA(nca_summary$PPORRES))

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = as.data.frame(nca_summary),
  reference     = published,
  by            = "wtgrp",
  units         = c(cmax = "ng/mL", aucinf.obs = "ng*h/mL", auclast = "ng*h/mL",
                    cl.obs = "L/h", tmax = "min", vss.obs = "L", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated (200 virtual adolescents per weight band) vs published Table 1 of Klein 2026. * marks a difference of more than 20%."
)
```

```{r nca-narrative, echo = FALSE, results = "asis"}
all_pt <- nca_scaled |>
  group_by(PPTESTCD) |>
  summarise(geomean = gm(PPORRES), median = median(PPORRES, na.rm = TRUE),
            .groups = "drop")
gv  <- function(p) all_pt$geomean[all_pt$PPTESTCD == p]
mv  <- function(p) all_pt$median[all_pt$PPTESTCD == p]
pct <- function(x, ref) sprintf("%+.1f%%", 100 * (x / ref - 1))

cat(sprintf(
"Pooled across both weight bands (the `All patients` column of Table 1, N = 14),
the simulated geometric means are Cmax %.1f ng/mL against a published 35.50
(%s), AUCinf %.0f ng*h/mL against 280.0 (%s), CL/F %.2f L/h against 7.144 (%s),
terminal half-life %.2f h against 7.601 (%s) and Vss/F %.1f L against 76.82
(%s); median Tmax is %.1f min against a published median of 10.50 min.\n\n",
gv("cmax"), pct(gv("cmax"), 35.50),
gv("aucinf.obs"), pct(gv("aucinf.obs"), 280.0),
gv("cl.obs"), pct(gv("cl.obs"), 7.144),
gv("half.life"), pct(gv("half.life"), 7.601),
gv("vss.obs"), pct(gv("vss.obs"), 76.82),
mv("tmax")))
```

```{r nca-assert}
# Structural agreement on the quantities that are not sampling-schedule limited.
# The published comparators are geometric means over 12-14 patients whose own
# standard errors are 8-16% (geometric CVs of 30-58% over sqrt(n)), so the bounds
# below are set against that sampling error, not tightened to the point estimate.
stopifnot(
  abs(gv("half.life")   / 7.601 - 1) < 0.10,   # terminal half-life
  abs(gv("aucinf.obs")  / 280.0 - 1) < 0.20,   # total exposure
  abs(gv("cl.obs")      / 7.144 - 1) < 0.20,   # apparent clearance
  abs(gv("cmax")        / 35.50 - 1) < 0.30    # peak; published SE alone is ~15%
)
# The simulated CL/F must sit near the model's own typical value; the published
# adolescent geometric mean is a 12-subject sample from a model fitted mostly to
# adults, so it is the noisier of the two comparators.
stopifnot(abs(gv("cl.obs") / cl_typ - 1) < 0.15)
# Median Tmax must land on a protocol sampling time near the published median.
stopifnot(mv("tmax") %in% round(protocol_times * 60, 6), abs(mv("tmax") - 10.5) <= 10)
```

```{r nca-flagged, echo = FALSE, results = "asis"}
flagged <- cmp[grepl("\\*", cmp$`% diff`), ]
if (nrow(flagged)) {
  cat("Rows flagged at the 20% tolerance: ",
      paste0("`", flagged[[1]], "` (", flagged$wtgrp, ")", collapse = ", "),
      ".\n", sep = "")
} else {
  cat("No row exceeds the 20% tolerance.\n")
}
```

Every flagged row is explained by the sampling design or by an NCA-estimator
definition rather than by the structural model, and none is a reason to adjust a
parameter:

* **`AUC0-t`.** The published `AUC0-t` is 99.4% of the published `AUCinf`, i.e.
  essentially every real patient stayed quantifiable to the final 36 h sample.
  The model's clearance IIV of 58.6% is estimated overwhelmingly from the 85
  adults, and it generates a heavier upper tail of clearance than the 14
  adolescents happened to show; those fast-clearing virtual subjects fall below
  the 0.200 ng/mL LLOQ before 36 h, which truncates their `AUC0-t` window. The
  `AUCinf` comparison, which extrapolates past the last quantifiable sample and
  is therefore insensitive to where censoring begins, agrees to within
  `r sprintf("%.0f%%", abs(100 * (gv("aucinf.obs") / 280.0 - 1)))`.
* **`Tmax`.** The published lower-band median of 19.50 min comes from six
  patients whose individual values span 2-120 min, so it is barely identified.
  Pooled over all 14 patients the published median is 10.50 min, which the
  simulation reproduces exactly.
* **`Vss/F`.** `PKNCA`'s `vss.obs` is `MRT * CL`, and for extravascular data
  `MRT` absorbs the mean absorption time as well as the disposition mean
  residence time. The paper's Phoenix-derived `Vss/F` is subject to the same
  effect but computed over a far sparser sampling grid, so the two estimators are
  not measuring quite the same quantity. The *ratio* between weight bands is the
  part the model actually predicts, and it agrees: the published 87.38 / 67.53 =
  `r sprintf("%.2f", 87.38 / 67.53)` against an allometric expectation of
  `r sprintf("%.2f", (64.35 / 41.27)^0.470)`.

`Cmax` runs about 18% below the published geometric mean without being flagged.
That gap is within the sampling error of the comparator: the published `Cmax`
has a geometric CV of 57.8% over 14 patients, i.e. a standard error on the
geometric mean of roughly 15%. The paper additionally warns that too few samples
were drawn during absorption (2, 10 and 30 min) to pin `Cmax` and `Tmax`
precisely, which biases a sparse-grid NCA `Cmax` in either direction depending on
where each subject's true peak falls between samples.

# Replicate Figure 4

Figure 4 plots simulated AUCinf and Cmax against body weight across the
adolescent range, with and without an enzyme-inducing antiseizure medication,
against the 90% range of adult values. AUCinf is reproduced in closed form
(`dose / CL`, as the Supplementary Methods specify); Cmax requires a solve, so a
weight-spanning cohort is simulated on a grid fine enough to resolve the peak.

```{r figure-4}
set.seed(20260829)
n_span <- 200

span <- tibble::tibble(
  id           = seq_len(2 * n_span),
  WT           = rep(runif(n_span, 25, 115), times = 2),
  CONMED_EIAED = rep(c(0, 1), each = n_span),
  panel        = rep(c("Without inducer ASMs", "With inducer ASMs"), each = n_span)
)

peak_times <- sort(unique(c(seq(0, 1, by = 1 / 120), seq(1, 6, by = 0.05))))

ev_span <- bind_rows(
  span |> mutate(time = 0, amt = 2, evid = 1L, cmt = "depot"),
  span |> mutate(time = 0, amt = 2, evid = 1L, cmt = "central"),
  tidyr::crossing(span, time = peak_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
) |>
  arrange(id, time, desc(evid))

sim_span <- rxode2::rxSolve(mod, ev_span, keep = c("WT", "panel")) |>
  as.data.frame()

per_subject <- sim_span |>
  group_by(id, WT, panel) |>
  summarise(cmax = max(Cc), cl = first(cl), .groups = "drop") |>
  mutate(aucinf = dose_ug / cl,
         wtbin  = cut(WT, breaks = seq(25, 115, by = 10)))

fig4 <- per_subject |>
  pivot_longer(c(aucinf, cmax), names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric, aucinf = "AUCinf (h.ng/mL)", cmax = "Cmax (ng/mL)")) |>
  group_by(panel, metric, wtbin) |>
  summarise(wt = mean(WT),
            p05 = quantile(value, 0.05), p50 = median(value),
            p95 = quantile(value, 0.95), .groups = "drop") |>
  mutate(panel = factor(panel, levels = c("Without inducer ASMs", "With inducer ASMs")))

ggplot(fig4, aes(wt, p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "firebrick", linewidth = 0.8) +
  facet_grid(metric ~ panel, scales = "free_y") +
  labs(x = "Body weight (kg)", y = NULL,
       title = "Figure 4 -- simulated exposure by body weight",
       caption = "Replicates Figure 4 of Klein 2026. Red: median; blue band: 90% range.")
```

Each claim below is asserted on a **per-subject regression across all 200
subjects in a panel**, not on the weight-binned medians that the figure plots.
A bin median is the median of roughly 22 log-normal draws and carries about 16%
standard error, so a bound tight enough to be meaningful on the underlying claim
would fail at random on a re-draw; the regression slope over the whole cohort
estimates the same quantity an order of magnitude more precisely.

```{r figure-4-assert}
ns  <- per_subject |> filter(panel == "Without inducer ASMs")
ind <- per_subject |> filter(panel == "With inducer ASMs")

# 1. AUCinf carries NO weight effect, because CL is not weight-scaled: the slope
#    of log(AUCinf) on log(WT) must be indistinguishable from zero.
slope_auc <- coef(stats::lm(log(aucinf) ~ log(WT), data = ns))[[2]]
stopifnot(abs(slope_auc) < 0.3)
stopifnot(abs(median(ns$aucinf) / (dose_ug / cl_typ) - 1) < 0.15)

# 2. Cmax must FALL with body weight at roughly the allometric exponent, because
#    Vc scales with weight while CL does not. Expected slope: -e_wt_vc_vp = -0.470.
slope_cmax <- coef(stats::lm(log(cmax) ~ log(WT), data = ns))[[2]]
stopifnot(slope_cmax > -0.9, slope_cmax < -0.1)

# 3. The inducer roughly halves AUCinf (the exact 1/1.955 ratio is asserted
#    deterministically in the structural-identity section above)...
stopifnot(abs((median(ind$aucinf) / median(ns$aucinf)) * 1.955 - 1) < 0.25)

# 4. ...and leaves Cmax essentially unchanged, as the paper states explicitly in
#    Results section 3.2: clearance barely affects the early peak.
stopifnot(abs(median(ind$cmax) / median(ns$cmax) - 1) < 0.20)
```

The simulation reproduces the three claims the authors draw from Figure 4:
AUCinf is flat across the adolescent weight range (no weight effect on
clearance); Cmax rises modestly at the low end of the weight range, driven by
the weight scaling of `Vc` and `Vp`; and co-administration of an enzyme-inducing
antiseizure medication roughly halves AUCinf while leaving Cmax essentially
unchanged.

# Assumptions and deviations

* **Reference weight for the allometric term (70 kg).** The paper reports a
  shared, freely estimated allometric exponent of 0.470 on `Vc` and `Vp` but
  never states the weight it is normalised to. 70 kg is assumed as the rounded
  standard. The assumption is corroborated rather than merely asserted: with
  `CL` unscaled, the allometric term is the only source of a weight effect on
  terminal half-life, and at 70 kg the model reproduces the observed Table 1
  half-lives in both adolescent weight bands to within
  `r sprintf("%.1f%%", max(abs(ident[["t1/2 error (%)"]])))`. A reference of
  75.5 kg (the pooled mean weight) fits the observed `Vss/F` slightly better but
  the half-lives distinctly worse; neither can be excluded from the published
  material.
* **Reading of the `IIV` column of Table S3.** Taken as `omega * 100` on each
  parameter's own estimation scale, not as the exact log-normal CV%, on the
  Figure 4 evidence set out above. The two readings differ by about 8% on omega.
* **Estimation scale of `F2` and of the inducer effect.** Neither is stated;
  both were recovered from the symmetry of their own published 95% CIs, as
  tabulated in the source-trace section. `F2` is carried on the logit scale so
  that individual dose fractions stay inside (0, 1).
* **Two dose records per administration.** The paper describes immediate release
  into the central compartment alongside first-order depot absorption but does
  not print its NONMEM dataset. The `F2` label (bioavailability of compartment 2
  in NONMEM's numbering, the central compartment for a two-compartment model with
  a depot) implies dose records into both compartments, which is how the model is
  encoded. An event table with a single dose record will silently deliver only
  one of the two absorption arms.
* **Covariates screened but not retained.** Height, sex, age, concomitant
  antiseizure medications other than the strong inducers, chronic benzodiazepine
  use, creatinine clearance and alanine aminotransferase were all screened in the
  stepwise covariate analysis and none was retained; no point estimate is
  published for any of them. They are recorded in the model's
  `covariatesDataExcluded` metadata rather than encoded. Body weight was also
  tested on `CL` and `Q` and rejected, so those two parameters are deliberately
  not weight-scaled.
* **Bioavailability is not identifiable.** Every disposition parameter is
  apparent (`/F`). `F2` partitions the apparent dose between the two absorption
  arms; it is not an absolute bioavailability.
* **Virtual cohort demographics.** Body weight is drawn from a normal
  distribution matched to the UP0100 mean and SD and truncated to the observed
  range; the paper publishes summary statistics only. Race, sex and age are not
  simulated because no retained covariate depends on them.
* **Adolescent population sizes.** The main text reports 84 patients enrolled in
  ENGAGE-E-001 while Table S1 reports 80 in the PK modelling dataset; the model's
  `population` metadata records the modelling dataset (N = 99 overall).
* No erratum or corrigendum was found for this article (CrossRef `update-to` and
  `updated-by` are both empty; a PubMed erratum search returns no records).
