---
title: "Pipamperone (Kloosterboer 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pipamperone (Kloosterboer 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Kloosterboer_2020_pipamperone"))$reference`
* Description: `r rxode2::rxode(readModelDb("Kloosterboer_2020_pipamperone"))$description`
* Article: <https://doi.org/10.1007/s40262-020-00894-y>
* PubMed Central: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7658071/>
* Electronic Supplementary Material (external-validation goodness-of-fit and
  NPDE plots only; no control stream):
  <https://static-content.springer.com/esm/art%3A10.1007%2Fs40262-020-00894-y/MediaObjects/40262_2020_894_MOESM1_ESM.pdf>

# Population

Kloosterboer 2020 built the model from 70 pipamperone concentrations in 30
children and adolescents: 8 patients from a Dutch multicentre observational
trial in autism spectrum disorder with comorbid behavioural problems (SPACe,
NTR6050; six samples per patient at random times on two days, by venepuncture
and dried blood spot, analysed by UHPLC-MS) and 22 patients from the
therapeutic-drug-monitoring service of the University Hospital Wuerzburg
(morning trough samples at steady state, analysed by HPLC-UV). Table 1 gives
the model-building group as 70% male, median age 13.0 years (range 5.6-17.7),
median body weight 50.4 kg (24.8-100.4), median BMI 20.4 kg/m^2, and a median
daily dose of 45 mg (12-400 mg) as tablets or oral solution given once to five
times daily. Diagnoses were autism spectrum disorder (73.3%), ADHD (40%),
mental retardation (30%), schizophrenia-spectrum disorders (6.7%) and conduct
disorder (3.3%); 60% took another antipsychotic concomitantly. A further 21
German TDM patients (33 concentrations) were used only for external
validation.

The same information is available programmatically:

```{r population}
str(rxode2::rxode(readModelDb("Kloosterboer_2020_pipamperone"))$population)
```

# Source trace

| Model element | Value | Source location |
|---|---|---|
| Structure: one compartment, first-order absorption, no lag | -- | Section 3.1.1 'best described using a one-compartment model'; Section 2.4.1 |
| `lka` (ka) | 2 /h, fixed | Table 2 'Ka' = 2, footnote a 'Fixed'; Section 2.4.1 'fixed at 2/h' |
| `lcl` (CL/F) | 22.1 L/h per 70 kg | Table 2 'CL/F (L/h/70 kg)' |
| `lvc` (V/F) | 416 L per 70 kg | Table 2 'V/F (L/70 kg)' |
| `e_wt_cl` | 0.75, fixed | Table 2 footnote b; Section 3.1.2 'fixed exponents' |
| `e_wt_vc` | 1, fixed | Table 2 footnote b |
| `etalcl` | 0.041153 (20.5% CV) | Table 2 'IPV CL' = 20.5%; `log(1 + 0.205^2)` |
| `addSd` | 0.21 ug/L | Table 2 'Additional error' |
| `propSd` | 0.39 | Table 2 'Proportional error' |
| `addSd_hplc` | 26.6 ug/L | Table 2 'Additional error HPLC-UV'; Section 3.1.3 |
| `cal_slope_dbs` | 0.33 | Table 2 'DBS correction: y = ax + b', a |
| `cal_int_dbs` | 3.90 ug/L | Table 2 'DBS correction: y = ax + b', b |
| Allometric scaling `(WT/70)^0.75` on CL, `(WT/70)^1` on V | -- | Section 2.4.1; Table 2 footnote b |
| DBS prediction `a * Cplasma + b` | -- | Table 2 (see Assumptions for the Discussion's form) |
| HPLC-UV extra additive error | -- | Section 3.1.1 'combined error model with an extra additional error for HPLC-UV concentrations' |

# Typical-value replication of Figure 4

Section 3.3 prints the population-predicted steady-state trough of a
twice-daily regimen for a 25, 50 and 75 kg patient, once for a fixed 30 mg dose
(Figure 4a) and once for 0.6 mg/kg (Figure 4b: 15, 30 and 45 mg). These are
typical values, so they are reproduced exactly with the random effects zeroed.

```{r regimens}
mod <- readModelDb("Kloosterboer_2020_pipamperone")

regimens <- tibble::tribble(
  ~treatment,           ~WT, ~dose, ~ctrough_pub,
  "30 mg BID, 25 kg",    25,    30,  163.2,
  "30 mg BID, 50 kg",    50,    30,  103.9,
  "30 mg BID, 75 kg",    75,    30,   79.3,
  "0.6 mg/kg BID, 25 kg", 25,   15,   81.6,
  "0.6 mg/kg BID, 50 kg", 50,   30,  103.9,
  "0.6 mg/kg BID, 75 kg", 75,   45,  119.0
) |>
  dplyr::mutate(id = dplyr::row_number())

obs_times <- sort(unique(c(seq(0, 12, by = 0.1), 12)))

make_events <- function(reg) {
  dose_rows <- reg |>
    dplyr::transmute(
      id, treatment, WT,
      time = 0, amt = dose, evid = 1, cmt = "depot", ss = 1, ii = 12
    )
  obs_rows <- reg |>
    dplyr::select(id, treatment, WT) |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(amt = 0, evid = 0, cmt = "central", ss = 0, ii = 0)
  dplyr::bind_rows(dose_rows, obs_rows) |>
    dplyr::mutate(ASSAY_LCMSMS = 1, SAMPLE_DBS = 0) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    as.data.frame()
}

ev_typ <- make_events(regimens)
sim_typ <- rxode2::rxSolve(rxode2::zeroRe(mod), events = ev_typ,
                           keep = c("treatment", "WT")) |>
  as.data.frame()
```

```{r fig4, fig.cap = "Replicates the population-prediction curves of Figure 4 of Kloosterboer 2020: steady-state twice-daily pipamperone, (a) fixed 30 mg, (b) 0.6 mg/kg."}
sim_typ |>
  dplyr::mutate(
    panel = ifelse(grepl("^30 mg", treatment), "(a) 30 mg BID", "(b) 0.6 mg/kg BID"),
    weight = paste(WT, "kg")
  ) |>
  ggplot(aes(time, Cc, colour = weight)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~panel) +
  labs(x = "Time after dose (h)", y = "Pipamperone (ug/L)", colour = "Body weight") +
  theme_bw()
```

The trough is also available in closed form for a one-compartment model with
first-order absorption at steady state, which checks the ODE solution
independently of rxode2:

```{r closed_form}
ss_trough <- function(wt, dose, tau = 12) {
  cl <- 22.1 * (wt / 70)^0.75
  v <- 416 * (wt / 70)
  ka <- 2
  k <- cl / v
  1000 * dose * ka / (v * (ka - k)) *
    (exp(-k * tau) / (1 - exp(-k * tau)) - exp(-ka * tau) / (1 - exp(-ka * tau)))
}

trough_chk <- sim_typ |>
  dplyr::filter(time == 12) |>
  dplyr::select(id, treatment, Cc) |>
  dplyr::left_join(regimens, by = c("id", "treatment")) |>
  dplyr::mutate(
    closed_form = ss_trough(WT, dose),
    pct_vs_paper = 100 * (Cc - ctrough_pub) / ctrough_pub
  )

trough_chk |>
  dplyr::select(treatment, Cc, closed_form, ctrough_pub, pct_vs_paper) |>
  dplyr::rename(
    "Regimen" = treatment,
    "rxode2 trough (ug/L)" = Cc,
    "Closed-form trough (ug/L)" = closed_form,
    "Kloosterboer 2020 Section 3.3 (ug/L)" = ctrough_pub,
    "% difference vs paper" = pct_vs_paper
  ) |>
  knitr::kable(digits = 2)

stopifnot(
  # Same drawn (typical) parameters on both sides: pure numerical error.
  all(abs(trough_chk$Cc - trough_chk$closed_form) / trough_chk$closed_form < 1e-3),
  # The paper prints the troughs to one decimal place.
  all(abs(trough_chk$pct_vs_paper) < 0.1)
)
```

All six published troughs are reproduced to the printed precision, which
confirms CL/F, V/F, the fixed ka, the 70 kg reference and both allometric
exponents together. The typical half-life at 70 kg is
`r round(log(2) * 416 / 22.1, 1)` h, matching the "mean elimination half-life
(13 h)" of the Discussion.

# PKNCA on the typical-value profiles

```{r pknca}
conc_df <- sim_typ |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

dose_df <- ev_typ |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | treatment + id,
                             concu = "ug/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "mg")
intervals <- data.frame(start = 0, end = 12, cmax = TRUE, tmax = TRUE,
                        auclast = TRUE, ctrough = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))

simulated_long <- as.data.frame(nca_res$result) |>
  dplyr::select(id, treatment, PPTESTCD, PPORRES)

published <- regimens |>
  dplyr::select(treatment, ctrough = ctrough_pub)

nlmixr2lib::ncaComparisonTable(
  simulated = simulated_long,
  reference = published,
  by = "treatment",
  units = c(cmax = "ug/L", auclast = "ug*h/L", ctrough = "ug/L", tmax = "h"),
  tolerance_pct = 20
) |>
  dplyr::rename("Regimen" = treatment) |>
  knitr::kable(caption = "Typical-value steady-state NCA over one 12 h interval versus the Section 3.3 population-predicted troughs.")
```

Kloosterboer 2020 reports only the troughs for these regimens; Cmax, Tmax and
AUC are shown for completeness. The 70 kg steady-state AUC over 24 h for a
60 mg daily dose is `r round(1000 * 60 / 22.1)` ug*h/L, in the range of the
median predicted AUC24h of the responders (3448 ug*h/L) and non-responders
(1811 ug*h/L) reported in Section 3.2.1 for the observed doses.

# Between-patient variability around the Figure 4 curves

Figure 4 shades a 95% interval from 1000 simulations around each curve. The
sketch below simulates 200 virtual patients per weight on the 0.6 mg/kg
regimen with between-patient variability on CL/F only; the residual error is
not added, so these are individual predictions.

```{r vpc}
rxode2::rxSetSeed(20200511)
n_per <- 200
reg_iiv <- regimens |>
  dplyr::filter(grepl("^0.6", treatment)) |>
  dplyr::select(treatment, WT, dose) |>
  tidyr::uncount(n_per) |>
  dplyr::mutate(id = dplyr::row_number(), ctrough_pub = NA_real_)

sim_iiv <- rxode2::rxSolve(mod, events = make_events(reg_iiv),
                           keep = c("treatment", "WT")) |>
  as.data.frame()

sim_iiv |>
  dplyr::group_by(treatment, time) |>
  dplyr::summarise(
    q025 = quantile(Cc, 0.025), q50 = median(Cc), q975 = quantile(Cc, 0.975),
    .groups = "drop"
  ) |>
  ggplot(aes(time, q50)) +
  geom_ribbon(aes(ymin = q025, ymax = q975), alpha = 0.25) +
  geom_line() +
  facet_wrap(~treatment) +
  labs(x = "Time after dose (h)", y = "Pipamperone (ug/L)") +
  theme_bw()

iiv_trough <- sim_iiv |>
  dplyr::filter(time == 12) |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    median = median(Cc), p2.5 = quantile(Cc, 0.025), p97.5 = quantile(Cc, 0.975),
    .groups = "drop"
  )
knitr::kable(iiv_trough, digits = 1)

typ_06 <- trough_chk |> dplyr::filter(grepl("^0.6", treatment))
chk <- dplyr::left_join(iiv_trough, typ_06, by = "treatment")
stopifnot(
  # Centre only: the median of a 200-patient cohort sits near the typical value.
  all(abs(chk$median - chk$Cc) / chk$Cc < 0.1)
)
```

The published intervals for 0.6 mg/kg (33.6-134.5, 49.0-167.2 and
57.6-186.7 ug/L at 25, 50 and 75 kg) are of similar width to, but somewhat
wider and lower-reaching than, the between-patient spread of CL/F alone
produces here. The paper does not state which variability components entered
its 1000 simulations (residual error at 39% proportional would widen them far
more than observed; parameter uncertainty would widen them modestly), so the
intervals are shown for orientation and are not used as a gate.

# Dried-blood-spot and HPLC-UV observations

A DBS observation row (`SAMPLE_DBS = 1`) returns the prediction on the DBS
scale, `0.33 * Cplasma + 3.90` ug/L. `ASSAY_LCMSMS = 0` (HPLC-UV) changes only
the residual error.

```{r dbs}
ev_dbs <- regimens |>
  dplyr::filter(treatment == "0.6 mg/kg BID, 50 kg") |>
  make_events() |>
  dplyr::filter(evid == 1 | time %in% c(2, 12))
ev_dbs2 <- ev_dbs |> dplyr::mutate(SAMPLE_DBS = ifelse(evid == 0, 1, 0))
cc_plasma <- rxode2::rxSolve(rxode2::zeroRe(mod), events = ev_dbs)$Cc
cc_dbs <- rxode2::rxSolve(rxode2::zeroRe(mod), events = ev_dbs2)$Cc
data.frame(time = c(2, 12), plasma = cc_plasma, dbs = cc_dbs) |>
  knitr::kable(digits = 2)
stopifnot(all(abs(cc_dbs - (0.33 * cc_plasma + 3.90)) < 1e-6))
```

# Assumptions and deviations

* **DBS conversion direction.** Table 2 prints the DBS correction as
  `y = ax + b` with a = 0.33 and b = 3.90 ug/L. The model reads this in the
  natural NONMEM form, `DBS prediction = a * plasma prediction + b`, which is
  the only reading of the table's equation that is consistent with the
  recovery of pipamperone in DBS being well below 100% (the earlier clinical
  validation divided DBS values by 0.158). The Discussion instead writes the
  conversion to plasma as `DBS/0.33 + 3.90`, which is equivalent to
  `DBS = 0.33 * (plasma - 3.90)`. The two readings share the slope and differ
  only in the intercept on the DBS scale (+3.90 versus -1.29 ug/L); the printed
  table equation is used. This affects DBS-scale predictions only, never the
  plasma model.
* **HPLC-UV residual error.** The paper describes "an extra additional error
  for HPLC-UV concentrations" without printing the error equation. It is added
  in variance to the common additive error, `sqrt(0.21^2 + 26.6^2)`; adding
  the standard deviations instead would give 26.81 rather than 26.60 ug/L, a
  difference of under 1%.
* **Residual-error form.** The combined error is encoded as nlmixr2's default
  (variances of the additive and proportional parts summed). Table 2 prints
  both residual terms as standard deviations.
* **IPV on CL/F.** The printed 20.5% is taken as a coefficient of variation of
  a log-normal random effect, `omega^2 = log(1 + 0.205^2)`.
* **ka unit.** Table 2 labels ka as 'L/h'; the Methods give it as '2/h', and
  1/h is used.
* **Assay coding.** The Dutch plasma and DBS methods are UHPLC-MS; they are
  coded `ASSAY_LCMSMS = 1`, and the German HPLC-UV method `ASSAY_LCMSMS = 0`.
* **Figure 4 intervals.** The 95% intervals of Figure 4 are not reproduced
  (see above); only the typical-value curves and troughs are.
