---
title: "Knibbe_2009_morphine"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Knibbe_2009_morphine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                     fig.width = 7, fig.height = 4.5,
                     warning = FALSE, message = FALSE)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r mod}
mod_obj <- rxode2::rxode2(readModelDb("Knibbe_2009_morphine"))
```

* Citation: `r mod_obj$reference`
* Description: `r mod_obj$description`
* Article: <https://doi.org/10.2165/00003088-200948060-00003> (PubMed [PMID 19650676](https://pubmed.ncbi.nlm.nih.gov/19650676/))
* DDMORE Foundation Model Repository entry:
  [DDMODEL00000248](https://repository.ddmore.eu/model/DDMODEL00000248)

This vignette validates the `Knibbe_2009_morphine` packaged model by
re-simulating the typical-value trajectories on the event table that
ships with the DDMORE bundle (`Simulated_PaediatricMorphinePK.csv`)
and visually comparing them to the bundle's stochastic-simulation
dependent variable. This is the F.2 self-consistency check from the
extraction skill's verification checklist (the source publication PDF
was not on disk during extraction, so a direct figure-replication
check against the publication is not possible).

# Population

Knibbe et al. 2009 fit a joint parent-metabolite population PK model for
morphine and its two glucuronide metabolites M3G and M6G in 248 preterm
neonates, term newborns, infants, and toddlers younger than 3 years. The
abstract reports 2,159 morphine concentrations across the 248 infants
and identifies bodyweight (BW exponent on clearance = 1.44) and a
postnatal-age step at PNA = 10 days as the structural covariates on the
glucuronidation pathways. The DDMORE-shipped `Output_real_run4.lst`
listing was fit to a `Combined_InternalExternalData.csv` dataset
covering 338 individuals / 2,809 observations / 5,302 records -- a
post-publication re-run on an extended pooled cohort. The full Knibbe
2009 publication PDF was not on disk during extraction; the model file's
`population` metadata flags the demographic detail fields as TODO.

The same metadata is available programmatically:

```{r population}
str(mod_obj$population, vec.len = 2, no.list = TRUE)
```

# Source trace

Per-parameter origin is recorded next to each `ini()` entry in
`inst/modeldb/ddmore/Knibbe_2009_morphine.R`. The table below collects
the structural-equation provenance in one place. All numeric values are
the FINAL PARAMETER ESTIMATE entries from the DDMORE-shipped
`Output_real_run4.lst` (status `MINIMIZATION SUCCESSFUL`, NSIG = 3.0);
the structural equations come from the `Executable_OriginalModelCode.mod`
control stream (`$MODEL`, `$PK`, `$ERROR` blocks).

| Equation / parameter                       | Value      | Source location                                                |
|--------------------------------------------|------------|----------------------------------------------------------------|
| `lvc` (V1 morphine, L per kg^allo_v)       | log(1.99)  | `Output_real_run4.lst` FINAL THETA TH 2 = 1.99E+00             |
| `lq` (Q morphine, L/min, no WT)            | log(0.0289)| `Output_real_run4.lst` FINAL THETA TH 3 = 2.89E-02             |
| `lfvm` (V_metab / V1, unitless)            | log(0.119) | `Output_real_run4.lst` FINAL THETA TH 8 = 1.19E-01             |
| `lcl_form_m3g_le10` (CL2, PNA <= 10 d)     | log(0.00309)| `Output_real_run4.lst` FINAL THETA TH 1 = 3.09E-03            |
| `lcl_form_m3g_gt10` (CL2, PNA > 10 d)      | log(0.00825)| `Output_real_run4.lst` FINAL THETA TH10 = 8.25E-03            |
| `lcl_form_m6g_le10` (CL3, PNA <= 10 d)     | log(0.000408)| `Output_real_run4.lst` FINAL THETA TH 6 = 4.08E-04          |
| `lcl_form_m6g_gt10` (CL3, PNA > 10 d)      | log(0.000699)| `Output_real_run4.lst` FINAL THETA TH11 = 6.99E-04          |
| `lcl_m3g` (M3G excretion CL4)              | log(0.00219)| `Output_real_run4.lst` FINAL THETA TH 7 = 2.19E-03            |
| `lcl_m6g` (M6G excretion CL5)              | log(0.00111)| `Output_real_run4.lst` FINAL THETA TH 9 = 1.11E-03            |
| `allo_cl` (BW exponent on CL)              | 1.44       | `Output_real_run4.lst` FINAL THETA TH 4 = 1.44E+00             |
| `allo_v` (BW exponent on V, FIXED)         | 1.00       | `Output_real_run4.lst` FINAL THETA TH 5 = 1.00E+00 (`1 FIX`)   |
| `etalcl_form_m3g`                          | 0.104      | `Output_real_run4.lst` FINAL OMEGA ETA1\|ETA1 = 1.04E-01       |
| `etalvc`                                   | 0.230      | `Output_real_run4.lst` FINAL OMEGA ETA2\|ETA2 = 2.30E-01       |
| `etalcl_m6g + etalcl_m3g ~ c(0.185, 0.178, 0.258)` | block | `Output_real_run4.lst` FINAL OMEGA BLOCK(2): ETA3\|ETA3, ETA4\|ETA3, ETA4\|ETA4 |
| `propSd` (morphine residual)               | sqrt(0.371)| `Output_real_run4.lst` FINAL SIGMA EPS1\|EPS1 = 3.71E-01       |
| `propSd_m3g`                               | sqrt(0.206)| `Output_real_run4.lst` FINAL SIGMA EPS2\|EPS2 = 2.06E-01       |
| `propSd_m6g`                               | sqrt(0.0967)| `Output_real_run4.lst` FINAL SIGMA EPS3\|EPS3 = 9.67E-02      |
| ODE for `central` (morphine, NONMEM CMT 1) | n/a        | `Executable_OriginalModelCode.mod` `$MODEL COMP (CENTRAL,DEFDOSE)`, `$PK` K12, K13, K14 |
| ODE for `peripheral1` (NONMEM CMT 4)       | n/a        | `$MODEL COMP (PERIPH)`, `$PK` K14, K41                         |
| ODE for `central_m3g` (NONMEM CMT 2)       | n/a        | `$MODEL COMP=(2)`, `$PK` K12, K20                              |
| ODE for `central_m6g` (NONMEM CMT 3)       | n/a        | `$MODEL COMP=(3)`, `$PK` K13, K30                              |
| Residual error (log-additive in NONMEM)    | n/a        | `$ERROR` block: `Y = LOG(F) + ERR(i)`, mapped to nlmixr2 proportional |

# Validation: F.2 self-consistency against the DDMORE simulated dataset

The DDMORE bundle ships `Simulated_PaediatricMorphinePK.csv`, an event
dataset for 5 subjects (BW 600 g-16,800 g, PNA 0.2 d-613 d) generated
by the same NONMEM model on stochastic IIV / EPS draws. We re-simulate
the same events with the packaged nlmixr2 model in **typical-value
mode** (`rxode2::zeroRe()` to drop the IIV / EPS contributions) and
overlay the typical-value trajectories on the bundle's per-record DV
values. The simulated DV is on the natural-log scale of ug/L
(equivalently ng/mL). Since the simulated DV carries IIV + residual
error and the typical-value trajectory does not, we expect the
typical-value line to pass through the cloud of DVs -- not to match
each value exactly.

```{r prepare-events}
sim_csv <- system.file(
  "extdata", "Knibbe_2009_morphine_Simulated_PaediatricMorphinePK.csv",
  package = "nlmixr2lib"
)
stopifnot(nzchar(sim_csv) && file.exists(sim_csv))

ds_raw <- read.table(sim_csv, header = TRUE, skip = 1)
events <- ds_raw |>
  dplyr::transmute(
    id   = ID,
    time = TIME,
    amt  = AMT,
    rate = RATE,
    evid = ifelse(MDV == 1, 1L, 0L),
    cmt  = dplyr::case_when(
      MDV == 1     ~ "central",      # all doses go to morphine central
      CMT == 1     ~ "Cc",
      CMT == 2     ~ "Cc_m3g",
      CMT == 3     ~ "Cc_m6g",
      TRUE         ~ NA_character_
    ),
    dv   = DV,                       # log(ug/L)
    cmt_label = dplyr::case_when(
      CMT == 1 ~ "morphine (Cc)",
      CMT == 2 ~ "M3G (Cc_m3g)",
      CMT == 3 ~ "M6G (Cc_m6g)"
    ),
    # NONMEM source units -> nlmixr2 canonical units
    WT  = BWS / 1000,                # grams -> kg
    PNA = PNA / 30.4375              # days  -> months
  )

# Add a uniform observation grid per subject so the typical-value trajectory
# is sampled on a fine resolution (the DDMORE event table only has ~3 obs
# per subject, which is too sparse to draw a smooth line).
grid_per_id <- events |>
  dplyr::group_by(id) |>
  dplyr::summarise(t_max = max(time), .groups = "drop") |>
  dplyr::rowwise() |>
  dplyr::reframe(
    id   = id,
    time = seq(0, t_max, length.out = 200),
    amt  = 0, rate = 0, evid = 0L,
    cmt_morph = "Cc", cmt_m3g = "Cc_m3g", cmt_m6g = "Cc_m6g"
  ) |>
  tidyr::pivot_longer(c(cmt_morph, cmt_m3g, cmt_m6g),
                      values_to = "cmt") |>
  dplyr::select(-name) |>
  dplyr::left_join(events |> dplyr::distinct(id, WT, PNA_init = PNA),
                   by = "id")
# PNA in the grid has to use a per-id PNA; in the source dataset PNA is
# time-varying but increases by ~1 day per 1440 min. Reconstruct PNA(t)
# linearly from the dataset's per-subject baseline PNA at t = 0 and the
# elapsed time in days.
grid_per_id <- grid_per_id |>
  dplyr::mutate(PNA = PNA_init + (time / 1440) / 30.4375) |>
  dplyr::select(id, time, amt, rate, evid, cmt, WT, PNA)

events_full <- dplyr::bind_rows(
  events |> dplyr::select(id, time, amt, rate, evid, cmt, WT, PNA),
  grid_per_id
) |>
  dplyr::arrange(id, time, evid)
```

```{r simulate-typical}
mod_typ <- rxode2::zeroRe(mod_obj)
sim_typ <- rxode2::rxSolve(mod_typ, events = events_full,
                           returnType = "data.frame")
head(sim_typ[, c("id", "time", "Cc", "Cc_m3g", "Cc_m6g")], 6)
```

```{r overlay-figure, fig.cap = "Typical-value rxode2 trajectories overlaid on the DDMORE-shipped Simulated_PaediatricMorphinePK.csv DV values, by analyte and subject. The DV column is on the natural-log scale (ug/L); the rxode2 prediction has been log-transformed to match. Horizontal axes show time in minutes; subjects are stratified by panel."}
typ_long <- sim_typ |>
  dplyr::transmute(id, time,
                   `morphine (Cc)`  = log(pmax(Cc,     1e-6)),
                   `M3G (Cc_m3g)`   = log(pmax(Cc_m3g, 1e-6)),
                   `M6G (Cc_m6g)`   = log(pmax(Cc_m6g, 1e-6))) |>
  tidyr::pivot_longer(-c(id, time), names_to = "analyte",
                      values_to = "log_conc")

obs_long <- events |>
  dplyr::filter(evid == 0) |>
  dplyr::transmute(id, time, analyte = cmt_label, log_conc = dv)

ggplot() +
  geom_line(data = typ_long,
            aes(time, log_conc, colour = analyte)) +
  geom_point(data = obs_long,
             aes(time, log_conc, colour = analyte),
             size = 2, alpha = 0.85) +
  facet_wrap(~ id, scales = "free", labeller = label_both) +
  labs(x = "Time (min)", y = "log(concentration in ug/L)",
       colour = NULL,
       title = "Knibbe 2009 -- typical-value rxode2 predictions vs DDMORE simulated DV") +
  theme_minimal()
```

# Quantitative agreement

Comparison of the typical-value rxode2 prediction at each observation
time-point against the DDMORE-shipped DV (which carries IIV + EPS).
Median and IQR of the residuals on the natural-log scale are reported
per analyte.

```{r residuals}
match_pred <- function(id_, time_, cmt_) {
  hit <- which(sim_typ$id == id_ & abs(sim_typ$time - time_) < 1e-6)
  if (length(hit) == 0) return(NA_real_)
  v <- sim_typ[[switch(cmt_, "Cc" = "Cc",
                              "Cc_m3g" = "Cc_m3g",
                              "Cc_m6g" = "Cc_m6g")]][hit[1]]
  log(pmax(v, 1e-6))
}

resid_tab <- events |>
  dplyr::filter(evid == 0) |>
  dplyr::rowwise() |>
  dplyr::mutate(pred_log = match_pred(id, time, cmt),
                resid    = dv - pred_log) |>
  dplyr::ungroup() |>
  dplyr::group_by(analyte = cmt_label) |>
  dplyr::summarise(
    n        = dplyr::n(),
    median   = round(stats::median(resid, na.rm = TRUE), 3),
    q25      = round(stats::quantile(resid, 0.25, na.rm = TRUE), 3),
    q75      = round(stats::quantile(resid, 0.75, na.rm = TRUE), 3),
    sd_obs   = round(stats::sd(resid, na.rm = TRUE), 3),
    .groups  = "drop"
  )
knitr::kable(resid_tab,
             caption = "Residuals (DV - typical-value PRED) on the log scale, by analyte. Median ~0 indicates structural fidelity; the IQR / SD reflects the IIV + residual variability sampled by NONMEM when the DDMORE simulated dataset was generated.")
```

For reference the source `.lst`'s reported FINAL OMEGA / SIGMA imply
expected log-scale SDs of approximately:

* morphine: sqrt(IIV2 on V1 + SIGMA1) = sqrt(0.230 + 0.371) ~= 0.78
* M3G: sqrt(IIV4 on CL4 + IIV1 on CL2 + SIGMA2) = sqrt(0.258 + 0.104 + 0.206) ~= 0.75
* M6G: sqrt(IIV3 on CL5 + SIGMA3) = sqrt(0.185 + 0.0967) ~= 0.53

The empirical residual SDs from the 5-subject DDMORE simulated event
table are within the same order as these analytic expectations -- small
deviations are expected because (a) the event table only includes
five randomly-drawn subjects, and (b) IIV on V1 also propagates into
the metabolite compartment volumes via `vc_m3g = fvm * vc`.

# Assumptions and deviations

- **Source publication PDF was not on disk during extraction.** All
  parameter values come from the DDMORE-shipped
  `Output_real_run4.lst` listing (`MINIMIZATION SUCCESSFUL`, NSIG=3.0).
  The publication abstract (PubMed PMID 19650676) confirms the headline
  numbers (n = 248 infants, BW exponent on CL = 1.44, PNA cutoff = 10
  days). Detailed demographic distributions, NCA tables, and figures
  from the publication could not be cross-checked. The model file's
  `population` metadata flags the affected fields as TODO.
- **Cohort size.** The publication's primary cohort is 248 infants /
  2,159 morphine concentrations; the DDMORE-shipped `.lst` is a
  `run4` re-fit on an extended pooled `Combined_InternalExternalData.csv`
  dataset with 338 individuals / 2,809 observations. The parameter
  values in this packaged model are the `run4` final estimates -- i.e.
  fit on the extended cohort, not on the publication's primary cohort.
- **Reference body weight.** The source NONMEM `.mod` uses
  `BW^exponent` (un-normalised) rather than `(BW / ref)^exponent`.
  The packaged model reproduces this verbatim; `units$dosing` and
  `units$concentration` are chosen so that the typical-value
  parameters match the `.lst` directly (implicit reference weight =
  1 kg, parameters are per-kg^exponent constants).
- **Postnatal-age threshold encoded in canonical months.** The source
  NONMEM `.mod` uses `IF (PNA.GT.10)` with PNA in days. The canonical
  PNA covariate (`inst/references/covariate-columns.md`) is in months,
  so the model uses `PNA > 10/30.4375` (~= 0.32855 months) for the
  switch. The arithmetic difference is exact (10/30.4375 days = 10
  days) but reviewers should note the unit-of-PNA difference between
  the source `.mod` and the packaged model.
- **Body weight time-varying input.** Source column `BWS` reports
  weight in grams; the packaged model expects `WT` in kg. The vignette
  performs the per-row conversion in `prepare-events`.
- **PNA-stratified clearance with shared IIV.** The source NONMEM
  `.mod` defines `CL2` (morphine -> M3G formation) with two THETAs
  (`THETA(1)` for PNA <= 10 d, `THETA(10)` for PNA > 10 d) and a single
  shared `ETA(1)` applied multiplicatively after the typical value is
  selected. The packaged model preserves this structure verbatim by
  exposing two ini parameters (`lcl_form_m3g_le10` and
  `lcl_form_m3g_gt10`) and one shared IIV (`etalcl_form_m3g`). This
  triggers a `checkModelConventions()` warning that the IIV
  `etalcl_form_m3g` has no matching fixed-effect parameter
  `lcl_form_m3g`; the deviation is intentional and documented here.
- **`$ERROR` block residual mapping.** The source `.mod` reports
  log-additive errors on log-IPRED (`Y = LOG(F) + ERR(i)`). Per
  `naming-conventions.md` `$ERROR block patterns`, this maps to
  `Cc ~ prop(propSd)` in nlmixr2 with `propSd = sqrt(SIGMA)`.
- **NKOD-conditional 24-hour additive error term dropped.** The
  source `.mod` includes a fourth `EPS(4)` slot
  (`+ TEH * ERR(4)` where `TEH = (TIME > 1900) & (NKOD in {1,2})`)
  that adds an extra additive variance for late post-infusion samples
  in one of the contributing studies. The gating column `NKOD` is
  study-specific to the `Combined_InternalExternalData.csv` dataset
  and is not part of the canonical-public model surface; it is
  dropped here. The corresponding `Output_real_run4.lst` SIGMA SIG4
  = 9.36E+00 is therefore not used.
- **Glucuronide molecular-weight conversion.** The source NONMEM
  `.mod` does not include a molecular-weight conversion factor when
  routing morphine mass-equivalents from `central` into the M3G /
  M6G compartments (morphine MW ~= 285, M3G MW ~= 461, M6G MW ~= 461).
  Whether the metabolite-compartment "concentration" is best
  interpreted as "M3G or M6G mass per L" or as "morphine
  mass-equivalents per L" depends on how the source dataset's DV
  column was assayed. The packaged model preserves the source
  parameterisation literally; users who need to interpret metabolite
  concentrations on a strict molar basis should multiply by
  MW(metab) / MW(morphine) when comparing simulated values to other
  PK studies.
```
