---
title: "BTZ-043 population PK and bacterial-load exposure-response (Koele 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BTZ-043 population PK and bacterial-load exposure-response (Koele 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
```

# Model and source

Koele 2025 contributed two models to nlmixr2lib, matching the two NONMEM runs
the paper reports:

* `Koele_2025_btz043` -- the joint BTZ-043 / M1 / M2 population PK model.
* `Koele_2025_btz043_bacterialload` -- the bilinear cfu + time-to-positivity
  exposure-response model, driven by the individual BTZ-043total AUC(0-24)
  that the PK model produces.

* Citation: `r rxode2::rxode(readModelDb("Koele_2025_btz043"))$reference`
* Article: <https://doi.org/10.1093/jac/dkaf076>

BTZ-043 is a first-in-class benzothiazinone inhibitor of DprE1, an enzyme of
the mycobacterial cell-wall synthesis pathway. The paper analyses the
sequential Phase 1b/2a trial NCT04044001, in which participants with
drug-susceptible pulmonary tuberculosis received BTZ-043 monotherapy at
250 to 1750 mg once daily for 14 days.

Both models were transcribed from the final NONMEM control streams printed in
the paper's Supplementary data, with typical values taken from Table 2 (PK)
and Table 3 (PD). Where the control stream and the tables disagree, the
resolution is recorded in the Errata section below.

```{r load-models}
mod_pk <- readModelDb("Koele_2025_btz043")
mod_pd <- readModelDb("Koele_2025_btz043_bacterialload")
ui_pk  <- rxode2::rxode(mod_pk)
ui_pd  <- rxode2::rxode(mod_pd)

# Molecular weight of BTZ-043, from the control stream comment on F1.
MW_BTZ043 <- 431.39
# 1 mg of BTZ-043 is this many nmol.
NMOL_PER_MG <- 1e6 / MW_BTZ043
# 1 nmol/L of BTZ-043 is this many ng/mL.
NGML_PER_NM <- MW_BTZ043 / 1000

c(nmol_per_mg = NMOL_PER_MG, ngmL_per_nM = NGML_PER_NM)
```

# Population

The analysis population is the 68 participants of NCT04044001 who received
BTZ-043 and contributed data: 24 from Stage 1 (Phase 1b dose escalation,
three participants per dose cohort and six in the highest) and 44 from
Stage 2 (Phase 2a, randomised 3:3:3:2 to 250, 500 or 1000 mg BTZ-043 daily
or to the Rifafour e-275 control regimen). One Stage 2 participant in the
500 mg group withdrew before any study procedure and contributed no data.

```{r population}
pop <- ui_pk$population
tibble::tibble(
  Field = names(pop),
  Value = vapply(pop, function(x) paste(paste0(
    ifelse(nzchar(names(x) %||% ""), paste0(names(x), ": "), ""), x),
    collapse = "; "), character(1))
) |>
  knitr::kable()
```

All participants were HIV-1 negative. Median age was 27 years (range 18 to 57),
median weight 54 kg (range 42 to 81) and 84% were male; 65% self-identified as
Black, 34% as Cape-coloured and 1% as White (Koele 2025 Table 1, combined
Stage 1 + 2 column). Note that the median weight of 54 kg sits well below the
70 kg allometric reference, so a typical participant in this trial has
apparent clearances about 18% lower than the tabulated 70 kg values.

Prandial state differed by stage and is a per-dose-record covariate rather
than a subject characteristic. Stage 1 dosed fasted on Days 1 to 12 and with
a high-fat breakfast on Day 14. Stage 2 dosed either immediately prior to, or
30 minutes after the start of, a standard breakfast; the latter is the model's
reference state.

# Source trace

Every `ini()` value in both model files carries an in-file comment naming its
source location. The table below is the consolidated audit trail.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Quantity,                                     ~Value,        ~Source,
  "Structural PK: CL/F, V/F, ka, MTT, Q/F, Vp/F", "404 L/h, 764 L, 1.69 /h, 0.340 h, 68.0 L/h, 382 L", "Table 2 'Structural parameters'; $THETA 1-4, 24, 25",
  "M1 disposition: CL_M1, V_M1",                 "36.0 L/h, 894 L", "Table 2; $THETA 13, 14",
  "M2 disposition: CL_M2, V_M2, Q_M2, Vp_M2",    "45.8 L/h, 19.0 L, 64.2 L/h, 26.8 L", "Table 2; $THETA 16, 17, 19, 20",
  "Parallel-absorption fraction, lag time",      "0.644, 1.83 h", "Table 2; $THETA 6, 7",
  "Allometric exponents",                        "0.75 CL, 1 V (fixed)", "Methods 'PK model development'; $PK AlloCL / AlloV",
  "F effects: dose >1250 mg, high fat, no food, dose before food", "0.710, 1.41, 0.458, 0.727", "Table 2 'Covariates'; $THETA 8-11",
  "MTT effect when no food is present",          "0.360",       "Table 2 'Dose prior to standard food/no food on MTT (%)'; $THETA 5",
  "M1 fm effect when no food is present",        "1.39 divisor", "Table 2 'Administration with food on FM1 (%)'; $THETA 23 under IF(WITHFOOD.EQ.0)",
  "Cape-coloured on CL/F",                       "0.762",       "Table 2; $THETA 22",
  "M2 clearance step after Day 10",              "0.734 from 240 h", "Table 2, Results; $THETA 21 under IF(DAY.GT.10)",
  "PK IIV and IOV variances",                    "see model file", "$OMEGA blocks; each reproduces the Table 2 CV% via CV = sqrt(exp(w2) - 1)",
  "PK residual errors",                          "50.5, 33.8, 29.0 CV%", "Table 2 'Residual error'",
  "PD baselines, Emax slopes, second slopes",    "6.20, 1.99, 0.0270, 0.00386, 0.00254, 0.000440", "Table 3; $THETA 1-6",
  "PD node, EC50",                               "48 h FIX, 16900 ng/mL*h", "Table 3; $THETA 7 FIX, 8",
  "PD IIV blocks and correlations",              "see model file", "$OMEGA BLOCK(3) and BLOCK(2); reproduce the Table 3 CV% and correlations",
  "PD residual errors",                          "0.558, 0.0624 log10", "Table 3 'Residual error' (published averages of the $SIGMA BLOCK(4) replicate SDs)"
) |>
  knitr::kable()
```

# Simulation helpers

Observation rows carry `cmt = "Cc"`. That is the ENDPOINT name, and for a
model with three declared endpoints it is the only form rxode2 accepts:
`cmt = "central"` (an ODE state) and `dvid`-only rows both fail with
`'dvid'->'cmt' ... on a undefined compartment`. Because `Cc` is a declared
endpoint (`Cc ~ prop(propSd)`), rxode2 already owns a compartment slot for it,
so nothing is injected and no ODE state is renumbered -- which the guard chunk
below proves rather than asserts in prose. `useLinCmt = FALSE` is set on every
solve so that rxode2's automatic ODE-to-linCmt conversion cannot silently
replace the transcribed ODE system.

```{r helpers}
# Trapezoidal AUC over an ordered time / concentration pair.
trap_auc <- function(time, conc) {
  sum(diff(time) * (utils::head(conc, -1) + utils::tail(conc, -1)) / 2)
}

# Prandial states of the trial, as the two registered indicators that span
# them (see the FED / FASTED_STRICT covariate notes in the model file).
PRANDIAL <- tibble::tribble(
  ~arm,                  ~FED, ~FED_HIGHFAT, ~FASTED_STRICT,
  "standard, dose after",  1,            0,              0,
  "high-fat",              1,            1,              0,
  "dose before standard",  0,            0,              0,
  "fasted",                0,            0,              1
)

# Build an event table: n_dose daily doses of `dose` mg into BOTH absorption
# depots, plus an observation grid. The dose must be written to depot1 and
# depot2 separately; f() then splits it between the transit chain and the
# lag-time secondary route.
btz_events <- function(dose, obs_times, n_dose = 1, ii = 24, id = 1L) {
  dosing <- tidyr::expand_grid(
    id = id, cmt = c("depot1", "depot2")
  ) |>
    dplyr::mutate(
      time = 0, amt = dose, evid = 1L, ii = ii,
      addl = as.integer(n_dose - 1L)
    )
  obs <- tidyr::expand_grid(id = id, time = obs_times) |>
    dplyr::mutate(
      cmt = "Cc", amt = NA_real_, evid = 0L, ii = 0, addl = 0L
    )
  dplyr::bind_rows(dosing, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

# Attach the covariate columns a solve needs.
with_covs <- function(ev, WT = 70, arm = "standard, dose after",
                      DOSE_HIGH = 0, RACE_COLOURED = 0, OCC = 1) {
  p <- PRANDIAL[match(arm, PRANDIAL$arm), ]
  stopifnot(nrow(p) == 1L, !is.na(p$FED))
  ev |>
    dplyr::mutate(
      WT = WT, FED = p$FED, FED_HIGHFAT = p$FED_HIGHFAT,
      FASTED_STRICT = p$FASTED_STRICT,
      DOSE_HIGH = DOSE_HIGH, RACE_COLOURED = RACE_COLOURED, OCC = OCC
    )
}

solve_pk <- function(ev, typical = TRUE, ...) {
  m <- if (typical) rxode2::zeroRe(mod_pk) else mod_pk
  rxode2::rxSolve(m, ev, returnType = "data.frame", useLinCmt = FALSE, ...)
}
```

# Structural verification: solve against closed form

The single most informative check on a transcribed multi-compartment model is
whether the solved system reproduces the closed forms the parameters imply.
With every fraction metabolised fixed to 1, the model routes the FULL parent
elimination flux into both metabolite compartments, so all three analytes have
the same closed-form AUC(0-inf) shape, `dose / CL`, with the dose expressed in
nmol. This is a deterministic identity: a tight tolerance is correct here and
would catch a mis-transcribed clearance, a lost allometric term, or an
accidental linCmt substitution of the transcribed ODEs.

```{r closed-form}
obs_grid <- seq(0, 96, by = 0.02)
ev_cf <- btz_events(dose = 500, obs_times = obs_grid) |> with_covs(WT = 70)
sim_cf <- solve_pk(ev_cf) |> dplyr::filter(!is.na(Cc))

dose_nmol <- 500 * NMOL_PER_MG
cf <- tibble::tibble(
  Analyte = c("BTZ-043", "M1", "M2"),
  `CL (L/h)` = c(404, 36.0, 45.8),
  `AUC simulated (nmol*h/L)` = c(
    trap_auc(sim_cf$time, sim_cf$Cc),
    trap_auc(sim_cf$time, sim_cf$Cc_m1),
    trap_auc(sim_cf$time, sim_cf$Cc_m2)
  )
) |>
  dplyr::mutate(
    `AUC closed form (nmol*h/L)` = dose_nmol / `CL (L/h)`,
    Ratio = `AUC simulated (nmol*h/L)` / `AUC closed form (nmol*h/L)`
  )
knitr::kable(cf, digits = c(0, 1, 1, 1, 5))
```

BTZ-043 and M2 are complete within the 96 h window and match their closed
forms exactly. M1 is the slowest analyte (`V_M1 / CL_M1` = 24.8 h, so a
17.2 h terminal half-life) and about 2% of its AUC still lies beyond 96 h,
which is the whole of its shortfall.

```{r closed-form-gate}
stopifnot(
  # Deterministic identity, not a cohort statistic: BTZ-043 and M2 are
  # numerically complete in the window, so the tolerance is solver accuracy.
  abs(cf$Ratio[cf$Analyte == "BTZ-043"] - 1) < 1e-3,
  abs(cf$Ratio[cf$Analyte == "M2"] - 1) < 1e-3,
  # M1 is truncated by design; exp(-96 / 24.83) = 2.1% remains uncollected.
  cf$Ratio[cf$Analyte == "M1"] > 0.96,
  cf$Ratio[cf$Analyte == "M1"] < 1.00
)
```

## Guard: no compartment renumbering

`cmt = "Cc"` on the observation rows is flagged by the repository's static
vignette lint as the slot-renumbering anti-pattern. That advice is right for a
BARE algebraic observable; it is a false positive for a declared endpoint. The
guard below turns the concern into a test that would go red if renumbering
ever did occur.

```{r guard-renumbering}
stopifnot(
  # every ODE state still resolves to a column
  all(ui_pk$state %in% names(sim_cf)),
  # every endpoint plus the derived total is returned
  all(c("Cc", "Cc_m1", "Cc_m2", "Cc_total") %in% names(sim_cf)),
  # The transit-chain depot receives exactly f(depot1) = 0.644 * 500 mg
  # converted to nmol, at t = 0.
  abs(sim_cf$depot1[1] - 0.644 * 500 * NMOL_PER_MG) < 1e-6,
  # depot2 is still EMPTY at t = 0 -- its dose is held by the 1.83 h lag --
  # and receives the complementary 0.356 fraction once the lag expires. The
  # observed peak sits on the first grid point at or after 1.83 h, by which
  # time ka has already removed exp(-1.69 * 0.01) = 1.7% of it, so the ratio
  # is just below 1 rather than equal to it.
  sim_cf$depot2[1] == 0,
  all(sim_cf$depot2[sim_cf$time < 1.83] == 0),
  max(sim_cf$depot2) / (0.356 * 500 * NMOL_PER_MG) > 0.97,
  max(sim_cf$depot2) / (0.356 * 500 * NMOL_PER_MG) <= 1,
  # and the derived total is exactly parent + M2
  max(abs(sim_cf$Cc_total - (sim_cf$Cc + sim_cf$Cc_m2))) < 1e-9
)
```

# Absorption: the double peak and the food-dependent secondary route

Koele 2025 Results: "A double-peak phenomenon was observed for BTZ-043, M1 and
M2 individual concentration-time profiles. A first-order secondary delayed
absorption into the BTZ-043 central compartment was identified when the drug
was taken together with food, using a lag-time model." The secondary route
carries 1 - 0.644 = 35.6% of the bioavailable dose after a 1.83 h lag, and is
switched off entirely when the dose is not taken with food.

```{r double-peak, fig.height = 4}
prof <- lapply(PRANDIAL$arm, function(a) {
  ev <- btz_events(500, seq(0, 12, by = 0.05)) |> with_covs(WT = 54, arm = a)
  solve_pk(ev) |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::transmute(time, Cc, arm = a)
}) |> dplyr::bind_rows()

ggplot(prof, aes(time, Cc, colour = arm)) +
  geom_line(linewidth = 0.8) +
  labs(
    x = "Time after dose (h)", y = "BTZ-043 (nmol/L)", colour = "Prandial state",
    title = "Typical-value BTZ-043 after 500 mg at 54 kg",
    subtitle = "Replicates the structure of Koele 2025 Figure 1 and Figures S1-S3"
  ) +
  theme_bw()
```

The two fed arms show the characteristic shoulder at about 2 to 3 h produced
by the delayed parallel route; the two arms with no food at the moment of
dosing rise and fall as a single peak, and peak earlier because the mean
transit time is 0.360 times the fed value.

```{r double-peak-gate}
peak_time <- prof |>
  dplyr::group_by(arm) |>
  dplyr::summarise(tmax = time[which.max(Cc)], .groups = "drop")
knitr::kable(peak_time, digits = 2)

fed_arms <- c("standard, dose after", "high-fat")
stopifnot(
  # Deterministic typical-value profiles, so exact ordering is safe here:
  # the fed arms peak on the delayed route, the unfed arms on the transit
  # chain with a 2.8-fold shorter mean transit time.
  all(peak_time$tmax[peak_time$arm %in% fed_arms] > 1.83),
  all(peak_time$tmax[!peak_time$arm %in% fed_arms] < 1.83)
)
```

# Prandial-state and dose effects on exposure

Because bioavailability enters only through `f()`, and none of the prandial
covariates touches a clearance, every published F multiplier must reappear
EXACTLY as a ratio of parent AUC(0-inf). The M1 ratios additionally carry the
1.39 divisor on apparent M1 clearance, which is the effect the paper describes
as a change in the relative fraction metabolised to M1.

```{r food-ratios}
auc_arm <- function(arm, WT = 54, dose = 500, DOSE_HIGH = 0,
                    RACE_COLOURED = 0) {
  ev <- btz_events(dose, seq(0, 120, by = 0.02)) |>
    with_covs(WT = WT, arm = arm, DOSE_HIGH = DOSE_HIGH,
              RACE_COLOURED = RACE_COLOURED)
  s <- solve_pk(ev) |> dplyr::filter(!is.na(Cc))
  c(
    parent = trap_auc(s$time, s$Cc),
    m1     = trap_auc(s$time, s$Cc_m1),
    m2     = trap_auc(s$time, s$Cc_m2)
  )
}

ref <- auc_arm("standard, dose after")
food_tab <- lapply(PRANDIAL$arm, function(a) {
  v <- auc_arm(a)
  tibble::tibble(
    arm = a,
    `BTZ-043 ratio` = unname(v["parent"] / ref["parent"]),
    `M1 ratio`      = unname(v["m1"] / ref["m1"]),
    `M2 ratio`      = unname(v["m2"] / ref["m2"])
  )
}) |>
  dplyr::bind_rows() |>
  dplyr::mutate(
    `Published F factor` = c(1, 1.41, 0.727, 0.458),
    `Published F x fm_M1` = `Published F factor` * c(1, 1, 1.39, 1.39)
  )
knitr::kable(food_tab, digits = 4)
```

```{r food-ratios-gate}
stopifnot(
  # Deterministic identities: AUC = F * dose / CL and no prandial covariate
  # touches CL, so the parent ratios must equal the published F factors.
  max(abs(food_tab$`BTZ-043 ratio` - food_tab$`Published F factor`)) < 1e-3,
  max(abs(food_tab$`M2 ratio`      - food_tab$`Published F factor`)) < 1e-3,
  # M1 additionally carries the 1.39 divisor on its apparent clearance. The
  # tolerance is looser only because M1 is the truncated analyte.
  max(abs(food_tab$`M1 ratio` - food_tab$`Published F x fm_M1`)) < 0.02
)
```

The M1 column is the one place where the paper's prose and its own control
stream point in opposite directions, and it is worth reading the numbers
rather than the sentence: in the model AS FITTED, M1 exposure is 39% HIGHER in
the arms with no food present at dosing. See the Errata section.

Two further deterministic effects reproduce exactly.

```{r dose-race-gate}
extra <- tibble::tibble(
  Effect = c("Dose > 1250 mg on F", "Cape-coloured on CL/F"),
  Simulated = c(
    unname(auc_arm("fasted", dose = 1500, DOSE_HIGH = 1)["parent"] /
             (1.5 * auc_arm("fasted", dose = 1000)["parent"])),
    unname(auc_arm("standard, dose after", RACE_COLOURED = 1)["parent"] /
             ref["parent"])
  ),
  Published = c(0.710, 1 / 0.762)
)
knitr::kable(extra, digits = 4)
stopifnot(max(abs(extra$Simulated - extra$Published)) < 1e-3)
```

# The M2 clearance step after 10 days on treatment

Koele 2025 Results: "The clearance of M2 decreased by 27% (95% CI: 24%-29%)
after 10 days on BTZ-043 treatment. The effect was implemented as a
dichotomous step after 10 days". Simulating 14 daily doses and comparing the
M2 AUC over the Day 9 interval with the Day 14 interval recovers the step.

```{r m2-step}
ev_ss <- btz_events(500, seq(0, 336, by = 0.05), n_dose = 14) |>
  with_covs(WT = 54)
sim_ss <- solve_pk(ev_ss) |> dplyr::filter(!is.na(Cc))

interval_auc <- function(s, from, to, col) {
  w <- s[s$time >= from & s$time <= to, ]
  trap_auc(w$time, w[[col]])
}
m2_day9  <- interval_auc(sim_ss, 192, 216, "Cc_m2")
m2_day14 <- interval_auc(sim_ss, 312, 336, "Cc_m2")
p_day9   <- interval_auc(sim_ss, 192, 216, "Cc")
p_day14  <- interval_auc(sim_ss, 312, 336, "Cc")

tibble::tibble(
  Analyte = c("M2", "BTZ-043"),
  `AUC(0-24) Day 9` = c(m2_day9, p_day9),
  `AUC(0-24) Day 14` = c(m2_day14, p_day14),
  `Day 14 / Day 9` = c(m2_day14 / m2_day9, p_day14 / p_day9),
  Expected = c(1 / 0.734, 1)
) |>
  knitr::kable(digits = 4)

stopifnot(
  # Deterministic typical-value steady state. M2 exposure rises by the
  # reciprocal of the published 73.4% factor; the parent is untouched.
  abs(m2_day14 / m2_day9 - 1 / 0.734) < 0.01,
  abs(p_day14 / p_day9 - 1) < 1e-3
)
```

```{r m2-step-plot, fig.height = 3.6}
sim_ss |>
  dplyr::filter(time >= 168) |>
  ggplot(aes(time / 24, Cc_m2)) +
  geom_line(linewidth = 0.6, colour = "#2b6cb0") +
  geom_vline(xintercept = 10, linetype = 2) +
  labs(
    x = "Day on treatment", y = "M2 (nmol/L)",
    title = "Typical-value M2 across the Day 10 clearance step",
    subtitle = "500 mg daily, 54 kg, standard breakfast; dashed line = 240 h"
  ) +
  theme_bw()
```

# Virtual cohort and PKNCA

Koele 2025 publishes no NCA table, so there is nothing to compare a
non-compartmental analysis against externally. PKNCA is used here for the
check that is available and is worth more anyway: an independent
implementation of AUC and Cmax run over a simulated cohort, with the
typical-value AUC(0-inf) compared against `F * dose / CL`.

```{r cohort}
rxode2::rxSetSeed(20250908)
n_per_arm <- 60L
arms <- c("standard, dose after", "high-fat", "fasted")

cohort_ev <- lapply(seq_along(arms), function(k) {
  ids <- seq_len(n_per_arm) + (k - 1L) * n_per_arm
  btz_events(500, sort(unique(c(
    0, seq(0.25, 12, by = 0.25), seq(13, 24, by = 1)
  ))), id = ids) |>
    with_covs(WT = 54, arm = arms[k]) |>
    dplyr::mutate(arm = arms[k])
}) |> dplyr::bind_rows()

sim_cohort <- rxode2::rxSolve(
  mod_pk, cohort_ev, returnType = "data.frame", useLinCmt = FALSE,
  keep = c("arm", "WT")
) |>
  dplyr::filter(!is.na(Cc))
if (is.null(sim_cohort$id)) sim_cohort$id <- 1L
nrow(sim_cohort)
```

```{r vpc, fig.height = 4.5}
vpc_dat <- sim_cohort |>
  dplyr::filter(arm == "standard, dose after") |>
  dplyr::select(id, time, `BTZ-043` = Cc, M1 = Cc_m1, M2 = Cc_m2) |>
  tidyr::pivot_longer(c(`BTZ-043`, M1, M2),
                      names_to = "analyte", values_to = "conc") |>
  dplyr::group_by(analyte, time) |>
  dplyr::summarise(
    lo  = quantile(conc, 0.025),
    mid = quantile(conc, 0.5),
    hi  = quantile(conc, 0.975),
    .groups = "drop"
  )

ggplot(vpc_dat, aes(time, mid)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.25, fill = "#2b6cb0") +
  geom_line(linewidth = 0.8, colour = "#2b6cb0") +
  facet_wrap(~analyte, scales = "free_y") +
  labs(
    x = "Time after dose (h)", y = "Concentration (nmol/L)",
    title = "Simulated 2.5th / 50th / 97.5th percentiles, 60 subjects",
    subtitle = "Replicates the layout of Koele 2025 Figure 2 (500 mg, standard breakfast, 54 kg)"
  ) +
  theme_bw()
```

```{r pknca}
conc_dat <- sim_cohort |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::transmute(id = as.integer(id), time, conc = Cc, treatment = arm)
stopifnot(nrow(conc_dat) > 0, all(conc_dat$conc >= 0))

dose_dat <- conc_dat |>
  dplyr::distinct(id, treatment) |>
  dplyr::mutate(time = 0, dose = 500)

o_conc <- PKNCA::PKNCAconc(conc_dat, conc ~ time | treatment + id)
o_dose <- PKNCA::PKNCAdose(dose_dat, dose ~ time | treatment + id)
o_data <- PKNCA::PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(
    start = 0, end = 24,
    cmax = TRUE, tmax = TRUE, auclast = TRUE, half.life = TRUE
  )
)
res <- suppressWarnings(PKNCA::pk.nca(o_data))
nca <- as.data.frame(res)

nca_summary <- nca |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "half.life")) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(
    Median = median(PPORRES, na.rm = TRUE),
    `P5` = quantile(PPORRES, 0.05, na.rm = TRUE),
    `P95` = quantile(PPORRES, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::rename("Prandial state" = treatment, "NCA parameter" = PPTESTCD)
knitr::kable(nca_summary, digits = 2)
```

The cohort medians order as the model requires, with the high-fat arm above
the standard-breakfast reference and the fasted arm below it.

```{r pknca-gate}
med_auc <- nca_summary |>
  dplyr::filter(`NCA parameter` == "auclast") |>
  dplyr::select(`Prandial state`, Median) |>
  tibble::deframe()
stopifnot(all(arms %in% names(med_auc)))

ratio_hf <- unname(med_auc["high-fat"] / med_auc["standard, dose after"])
ratio_fa <- unname(med_auc["fasted"] / med_auc["standard, dose after"])
c(high_fat = ratio_hf, fasted = ratio_fa)

stopifnot(
  # These are COHORT medians, not the deterministic ratios checked earlier,
  # so the bounds admit the sampling noise of a 60-subject arm while still
  # going red on a mis-transcribed F factor (which would move them by tens
  # of percent). Published factors are 1.41 and 0.458.
  ratio_hf > 1.15, ratio_hf < 1.75,
  ratio_fa > 0.33, ratio_fa < 0.62
)
```

A separate deterministic check ties PKNCA's own integration back to the
closed form, on the typical-value profile over a window long enough for
BTZ-043 to be numerically complete.

```{r pknca-closed-form}
ev_tv <- btz_events(500, sort(unique(c(
  seq(0, 12, by = 0.05), seq(12.5, 72, by = 0.5)
)))) |> with_covs(WT = 70)
sim_tv <- solve_pk(ev_tv) |> dplyr::filter(!is.na(Cc))

tv_conc <- data.frame(id = 1L, treatment = "typical",
                      time = sim_tv$time, conc = sim_tv$Cc)
tv_dose <- data.frame(id = 1L, treatment = "typical", time = 0, dose = 500)
tv_res <- suppressWarnings(PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(tv_conc, conc ~ time | treatment + id),
  PKNCA::PKNCAdose(tv_dose, dose ~ time | treatment + id),
  intervals = data.frame(start = 0, end = Inf, aucinf.obs = TRUE, cmax = TRUE)
)))
tv_nca <- as.data.frame(tv_res)
aucinf <- tv_nca$PPORRES[tv_nca$PPTESTCD == "aucinf.obs"]
stopifnot(length(aucinf) == 1L, !is.na(aucinf))

tibble::tibble(
  Quantity = "BTZ-043 AUC(0-inf), 500 mg, 70 kg, standard breakfast",
  `PKNCA (nmol*h/L)` = aucinf,
  `Closed form dose/CL (nmol*h/L)` = dose_nmol / 404,
  Ratio = aucinf / (dose_nmol / 404)
) |>
  knitr::kable(digits = c(0, 1, 1, 4))

stopifnot(
  # Deterministic: PKNCA log-linear extrapolation of a typical-value profile
  # against the model's own closed form. Tolerance is the trapezoidal error
  # of the 0.05 h / 0.5 h grid.
  abs(aucinf / (dose_nmol / 404) - 1) < 0.01
)
```

# Exposure-response: linking the two models

The PD model is driven by `AUC_BTZ043TOT`, the individual BTZ-043total
(BTZ-043 + M2) AUC over the 24 h dosing interval expressed in ng/mL*h. The PK
model works in nmol/L, so the link between the two models is one multiplication
by `MW / 1000`. The paper computed the metric on Day 12 (Stage 1) or Day 14
(Stage 2); Day 14 is used here, which is on the far side of the M2 clearance
step.

```{r auc-link}
btz_total_auc <- function(dose, WT = 54, arm = "standard, dose after",
                          DOSE_HIGH = 0) {
  ev <- btz_events(dose, seq(312, 336, by = 0.02), n_dose = 14) |>
    with_covs(WT = WT, arm = arm, DOSE_HIGH = DOSE_HIGH, OCC = 3)
  s <- solve_pk(ev) |> dplyr::filter(!is.na(Cc))
  trap_auc(s$time, s$Cc_total) * NGML_PER_NM
}

EC50 <- exp(ui_pd$theta[["lec50"]])
auc500 <- btz_total_auc(500)
tibble::tibble(
  Quantity = c("Simulated BTZ-043total AUC(0-24), Day 14, 500 mg, standard breakfast, 54 kg",
               "Published EC50"),
  `ng/mL*h` = c(auc500, EC50)
) |>
  knitr::kable(digits = 0)
```

Koele 2025 Results state the estimated EC50 is "similar to the exposure
obtained after a 500 mg BTZ-043 dose following a standard breakfast", and the
Discussion repeats it as support for the 500 mg arm of the DECISION trial.
The reconstruction puts that exposure at `r round(auc500)` ng/mL*h against a
published EC50 of `r round(EC50)` ng/mL*h -- the same quantity to within about
`r round(100 * abs(auc500 / EC50 - 1))`%, which is what a qualitative "similar
to" supports given that the paper's metric is a post-hoc individual estimate
averaged over a cohort whose weights and prandial states varied.

```{r ec50-gate}
stopifnot(
  # A qualitative claim ("similar to"), reconstructed deterministically. The
  # gate is a factor of two either way: it goes red on a wrong molar-to-mass
  # conversion (a 431-fold error), a wrong analyte (parent alone is 13-fold
  # below the total), or a wrong dosing day, but does not pretend the paper
  # stated a number it did not.
  auc500 / EC50 > 0.5, auc500 / EC50 < 2
)
```

## The published dose-response claims

The abstract's headline finding is that "Participants in the highest dose
group in Stage 2 (1000 mg) had a 2-fold faster decrease in mycobacterial load
during the initial 2 days compared with participants in the lowest dose group
(250 mg)". The Discussion adds that "The bacterial load decreased 5-6-fold
faster in the first 2 days on treatment compared with Days 2-14". Both are
consequences of the Emax first-phase slope and the exposure-independent
second-phase slope, so both can be reconstructed deterministically from the
two packaged models chained together.

```{r dose-response}
stage2_doses <- c(250, 500, 1000)
auc_by_dose <- vapply(stage2_doses, btz_total_auc, numeric(1))

emax_cfu   <- exp(ui_pd$theta[["lemax_cfu"]])
slope2_cfu <- exp(ui_pd$theta[["lslope2_cfu"]])
slope1_cfu <- emax_cfu * auc_by_dose / (EC50 + auc_by_dose)

dr <- tibble::tibble(
  `Dose (mg)` = stage2_doses,
  `BTZ-043total AUC(0-24) (ng/mL*h)` = auc_by_dose,
  `Slope 1 cfu (log10/h)` = slope1_cfu,
  `Slope 1 / slope 2` = slope1_cfu / slope2_cfu
)
knitr::kable(dr, digits = c(0, 0, 5, 2))

ratio_1000_250 <- slope1_cfu[stage2_doses == 1000] /
  slope1_cfu[stage2_doses == 250]
c(`slope1 1000mg / slope1 250mg` = ratio_1000_250)
```

```{r dose-response-gate}
stopifnot(
  # "2-fold faster" (abstract and Discussion). Deterministic given the two
  # models, so the band is narrow; it goes red if the EC50, either Emax or
  # the molar-to-mass conversion is mis-transcribed.
  ratio_1000_250 > 1.7, ratio_1000_250 < 2.3,
  # "5-6-fold faster in the first 2 days" (Discussion). The paper's figure is
  # a cohort average over the three Stage 2 dose groups, so the check is that
  # the reconstructed per-dose range brackets it rather than that any single
  # dose hits it.
  min(dr$`Slope 1 / slope 2`) < 5, max(dr$`Slope 1 / slope 2`) > 6
)
```

# Bacterial-load trajectories

The PD model has no ODE states and no dosing events: it is a bilinear
regression on time since the start of treatment, with a shared node fixed at
48 h. Both endpoints are on the log10 scale, cfu falling and TTP rising.

```{r pd-typical}
pd_events <- function(auc, times, id = 1L) {
  tidyr::expand_grid(id = id, time = times) |>
    dplyr::mutate(evid = 0L, AUC_BTZ043TOT = auc)
}

pd_times <- sort(unique(c(seq(0, 336, by = 2), 48)))
pd_typ <- lapply(seq_along(stage2_doses), function(k) {
  s <- rxode2::rxSolve(
    rxode2::zeroRe(mod_pd), pd_events(auc_by_dose[k], pd_times),
    returnType = "data.frame"
  )
  dplyr::transmute(s, time, log_cfu, log_ttp,
                   dose = paste0(stage2_doses[k], " mg"))
}) |> dplyr::bind_rows()

pd_typ |>
  tidyr::pivot_longer(c(log_cfu, log_ttp), names_to = "endpoint",
                      values_to = "value") |>
  dplyr::mutate(endpoint = dplyr::recode(
    endpoint,
    log_cfu = "log10 cfu/mL", log_ttp = "log10 TTP (h)"
  )) |>
  ggplot(aes(time / 24, value, colour = dose)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = 2, linetype = 2) +
  facet_wrap(~endpoint, scales = "free_y") +
  labs(
    x = "Day on treatment", y = NULL, colour = "BTZ-043 dose",
    title = "Typical-value bilinear bacterial-load decline",
    subtitle = "Dashed line = the node fixed at 48 h; Stage 2 dose groups"
  ) +
  theme_bw()
```

```{r pd-structural-gate}
base_row <- pd_typ |> dplyr::filter(time == 0, dose == "500 mg")
node_row <- pd_typ |> dplyr::filter(time == 48, dose == "500 mg")
stopifnot(nrow(base_row) == 1L, nrow(node_row) == 1L)

# Baselines at t = 0 must be the published typical values exactly.
stopifnot(
  abs(base_row$log_cfu - 6.20) < 1e-8,
  abs(base_row$log_ttp - 1.99) < 1e-8
)

# The bilinear form must be continuous at the node: the value there equals
# baseline minus (or plus) slope1 * 48 with no second-phase contribution.
auc500_lo <- auc_by_dose[stage2_doses == 500]
s1_cfu <- emax_cfu * auc500_lo / (EC50 + auc500_lo)
s1_ttp <- exp(ui_pd$theta[["lemax_ttp"]]) * auc500_lo / (EC50 + auc500_lo)
stopifnot(
  abs(node_row$log_cfu - (6.20 - s1_cfu * 48)) < 1e-8,
  abs(node_row$log_ttp - (1.99 + s1_ttp * 48)) < 1e-8
)

# And the second phase must be shallower than the first for cfu.
late <- pd_typ |> dplyr::filter(dose == "500 mg", time %in% c(48, 336))
stopifnot(diff(range(late$log_cfu)) / (336 - 48) < s1_cfu)
```

```{r pd-cohort, fig.height = 4.5}
rxode2::rxSetSeed(20250908)
n_pd <- 100L
pd_obs_times <- c(0, 48, 72, 96, 144, 192, 264, 336)

pd_cohort_ev <- pd_events(auc_by_dose[stage2_doses == 500],
                          pd_obs_times, id = seq_len(n_pd))
pd_cohort <- rxode2::rxSolve(mod_pd, pd_cohort_ev,
                             returnType = "data.frame")
if (is.null(pd_cohort$id)) pd_cohort$id <- 1L

pd_cohort |>
  dplyr::select(id, time, `log10 cfu/mL` = log_cfu,
                `log10 TTP (h)` = log_ttp) |>
  tidyr::pivot_longer(c(`log10 cfu/mL`, `log10 TTP (h)`),
                      names_to = "endpoint", values_to = "value") |>
  dplyr::group_by(endpoint, time) |>
  dplyr::summarise(
    lo = quantile(value, 0.025), mid = quantile(value, 0.5),
    hi = quantile(value, 0.975), .groups = "drop"
  ) |>
  ggplot(aes(time / 24, mid)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.25, fill = "#b7791f") +
  geom_line(linewidth = 0.8, colour = "#b7791f") +
  facet_wrap(~endpoint, scales = "free_y") +
  labs(
    x = "Day on treatment", y = NULL,
    title = "Simulated 2.5th / 50th / 97.5th percentiles, 100 subjects at 500 mg",
    subtitle = "Replicates the layout of Koele 2025 Figure 4 (without the censoring panels)"
  ) +
  theme_bw()
```

```{r pd-cohort-gate}
bl <- pd_cohort |> dplyr::filter(time == 0)
end <- pd_cohort |> dplyr::filter(time == 336)
stopifnot(nrow(bl) == n_pd, nrow(end) == n_pd)

# Cohort statistics, so these are magnitude and trend checks, not exact
# values. Published baseline CVs are 15.2% (cfu) and 5.14% (TTP); the bands
# below sit well outside a 100-subject sampling spread but still go red on a
# variance transcribed on the wrong scale (a variance-vs-SD confusion would
# move them several-fold).
cv <- function(x) sd(x) / mean(x)
c(cv_cfu = cv(bl$log_cfu), cv_ttp = cv(bl$log_ttp))
stopifnot(
  cv(bl$log_cfu) > 0.08, cv(bl$log_cfu) < 0.25,
  cv(bl$log_ttp) > 0.02, cv(bl$log_ttp) < 0.10,
  # Trend over 14 days: cfu falls, TTP rises, in the cohort median.
  median(end$log_cfu) < median(bl$log_cfu),
  median(end$log_ttp) > median(bl$log_ttp)
)
```

# Assumptions and deviations

* **Two dose records per administration.** The parallel-absorption structure
  needs the dose written to `depot1` and `depot2` at each dosing time; `f()`
  then splits it. This mirrors the source dataset, whose `DOSEX` column flags
  the second dose record for the parallel route. A simulation that doses only
  `depot1` silently drops 35.6% of the fed bioavailable dose.

* **Molar bookkeeping.** Amounts are nmol and concentrations nmol/L, following
  the source analysis ("Concentration data were transformed to molar units to
  preserve mass balance"). The mg-to-nmol conversion sits inside `f()`, so
  `units$dosing` is `"mg"` while `units$concentration` is `"nmol/L"`; the
  convention checker emits an informational note about this by design.

* **Fractions metabolised fixed to 1.** The full parent elimination flux feeds
  BOTH metabolite compartments, so metabolite mass balance is relative, not
  absolute, and every clearance and volume is an apparent value. This is the
  authors' choice, not a transcription artefact; changing it would invalidate
  every published estimate.

* **The M2 clearance step is derived from time, not from a data column.** The
  source dataset carries a `DAY` item and switches under `IF(DAY.GT.10)`. With
  Day 1 spanning 0 to 24 h this is `t >= 240 h`, which is how the model derives
  it, so no extra column is needed. A simulation whose event table does not
  start at the beginning of treatment will place the step incorrectly.

* **Prandial state is encoded as two registered indicators, not one
  four-level factor.** `FED` marks food already present at dosing and
  `FASTED_STRICT` marks a dose with no food on either side; `FED_HIGHFAT`
  refines the fed level. The fourth state -- dose taken immediately before a
  standard breakfast -- is the complement `(1 - FED) * (1 - FASTED_STRICT)`,
  derived inside `model()`. This reproduces the control stream's `FOOD` codes
  exactly.

* **Between-occasion variability is retained.** Relative bioavailability
  carries IIV plus IOV; mean transit time and ka carry IOV ONLY, as a
  correlated pair. Simulations in this vignette set `OCC = 1` throughout, i.e.
  a single occasion, which is what the covariate documentation recommends when
  the occasion structure of the source trial is not being reproduced.

* **Not encoded: censoring.** The PD model's M3 partial likelihood, its LLOQ
  of 1 cfu/mL and ULOQ of `log10(25 * 24)` for TTP, and the baseline
  probability of a negative culture (0.9% for cfu, 0% for TTP) have no
  idiomatic nlmixr2lib encoding and are omitted. The packaged model predicts
  the uncensored trajectory; the censoring panels of Koele 2025 Figure 4 are
  therefore not reproduced.

* **Not encoded: replicate-level and cross-endpoint residual correlation.**
  The PD `$SIGMA BLOCK(4)` resolves four replicate-level SDs with a full
  correlation matrix (see Errata). The packaged model uses the published
  per-endpoint averages with independent residuals.

* **Not encoded: the BLQ handling of the PK dataset.** The published PK model
  excluded the 600 (33%) BTZ-043, 159 (8.8%) M1 and 146 (8.1%) M2 observations
  below the 20 ng/mL limit of quantification, because the M3 re-estimation was
  unstable. Simulations from the packaged model produce concentrations
  throughout the elimination phase that the source analysis never fitted; the
  paper is explicit that the BTZ-043 late-elimination phase is the part the
  model describes least well.

* **No published NCA table.** Koele 2025 reports no Cmax / Tmax / AUC /
  half-life summary, so the PKNCA section compares against the model's own
  closed forms rather than against published non-compartmental values.

* **`cmt = "Cc"` on observation rows.** Required, not optional: with three
  declared endpoints rxode2 rejects an ODE-state `cmt` and rejects `dvid`-only
  rows. The repository's static vignette lint flags this form; the guard chunk
  above demonstrates that no compartment is renumbered.

# Errata and source conflicts

Three places where the paper is internally inconsistent. In each case the
resolution is recorded here and in the model file comments.

**1. The mean-transit-time sentence is inverted relative to the paper's own
table and control stream.** Koele 2025 Results says "The mean transit time
decreased by 64% (95% CI: 53%-72%) when taken together with any kind of food."
The corresponding Table 2 row is labelled "Dose prior to standard food/no food
on MTT (%) 36.0 (27.6-46.8)", and the control stream applies the factor under
`IF(WITHFOOD.EQ.0)MTTeff = THETA(5)`. Table and code agree with each other and
with the physiology (food delays gastric emptying, so MTT is LONGER with food);
the prose sentence has the direction backwards. The model follows the table and
the code: MTT is 0.340 h with food and 0.340 * 0.360 = 0.122 h without.

**2. The M1 fraction-metabolised effect is applied to the no-food arm in the
control stream, while the Table 2 row label and the prose attribute it to the
fed arm.** The control stream reads `FM1FOOD = 1; IF(WITHFOOD.EQ.0)FM1FOOD =
THETA(23)`, then `CLM1 = TVCLM1/FM1FOOD` and `VM1 = TVVM1/FM1FOOD`, so M1
exposure is 1.39-fold HIGHER when no food is present at dosing. Table 2 labels
the row "Administration with food on FM1 (%) 141 [sic 139]" and the Results
say the fraction metabolised to M1 "increased by 39% ... when BTZ-043 was
administered together with food". The model follows the executable control
stream, for three reasons: it is the artefact that produced the published
numbers; the immediately preceding MTT effect uses the identical
`IF(WITHFOOD.EQ.0)` gate and there the table label agrees with the code, so
the labelling convention in that block of Table 2 is "the arm the theta applies
to"; and the same paragraph of the Results contains the demonstrably inverted
MTT sentence. A user who needs the opposite direction can flip it by changing
the exponent on `e_fed_fm_m1` in `model()` from `(1 - FED)` to `FED`; the
absolute M1 concentrations move by a factor of 1.39 in the fed arms.

**3. Table 3 prints an impossible variability for the first TTP slope.** The
row reads "Emax slope 1 TTP ... 75.9 (35.4-67.1)": the point estimate lies
outside its own confidence interval, and 75.9 is exactly the value printed one
row below for the second cfu slope. The control stream's `$OMEGA BLOCK(3)`
third diagonal is 0.217, which gives CV = sqrt(exp(0.217) - 1) = 49.2% -- inside
the printed 35.4 to 67.1 interval. The model uses 0.217 and treats 75.9 as a
copy-paste error.

Two smaller discrepancies, resolved in favour of the published tables because
the supplementary `$THETA` blocks are initial estimates:

* The PK control stream's residual-error initials (0.477, 0.329, 0.285) differ
  from the Table 2 finals (0.505, 0.338, 0.290). Every other PK `$THETA` and
  every `$OMEGA` in that control stream equals the published final estimate
  exactly, which is what makes the three residual entries identifiable as
  stale initials. Table 2 is used.
* The PD control stream rounds two slopes (0.0039 and 0.0004) relative to the
  Table 3 finals (0.00386 and 0.000440). Table 3 is used.

Finally, the supplementary `$PRED` block omits `EXP(ETA(3))` from
`BETA1BTZTTP`, even though `ETA(3)` is declared in `$OMEGA BLOCK(3)` with a
non-zero variance and two estimated correlations, Table 3 reports its
variability, and the Results state that IIV was identified on the first TTP
slope. The packaged model applies it; the omission is a transcription slip.

For completeness, the full PD residual covariance the packaged model
simplifies:

```{r sigma-block, echo = FALSE}
sig <- matrix(c(
  0.299,   0.278,  -0.018,  -0.0175,
  0.278,   0.322,  -0.0194, -0.0189,
 -0.018,  -0.0194,  0.0043,  0.0033,
 -0.0175, -0.0189,  0.0033,  0.0035
), nrow = 4, byrow = TRUE,
dimnames = list(
  c("cfu rep 1", "cfu rep 2", "TTP rep 1", "TTP rep 2"),
  c("cfu rep 1", "cfu rep 2", "TTP rep 1", "TTP rep 2")
))
knitr::kable(sig, digits = 4, caption = "Supplementary $SIGMA BLOCK(4) (variances on the diagonal)")
sd_cfu <- mean(sqrt(diag(sig)[1:2]))
sd_ttp <- mean(sqrt(diag(sig)[3:4]))
tibble::tibble(
  Endpoint = c("cfu", "TTP"),
  `Mean replicate SD` = c(sd_cfu, sd_ttp),
  `Table 3 value` = c(0.558, 0.0624)
) |>
  knitr::kable(digits = 4)
stopifnot(abs(sd_cfu - 0.558) < 0.002, abs(sd_ttp - 0.0624) < 0.002)
```

# Session information

```{r session-info}
sessionInfo()
```
