---
title: "Apixaban (Kolowrat 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Apixaban (Kolowrat 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Kolowrat_2025_apixaban"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1111/cts.70392>
* Open access via Europe PMC: <https://europepmc.org/article/MED/PMC12597975>

Kolowrat 2025 asks a question that a conventional drug-drug interaction study
cannot answer. Amiodarone's half-life is about 50 days, so a crossover design in
healthy volunteers is neither ethical nor feasible; instead the authors built a
population PK model from *salvaged* plasma -- samples drawn for routine clinical
care that had passed a five-day discard threshold -- in 106 hospitalized patients
with nonvalvular atrial fibrillation, 51 of whom were on a stable dose of
amiodarone 200 mg.

The model is deliberately simple: one compartment, first-order absorption, no lag
time, and only two retained covariates. Two-compartment models were explored but
the peripheral volume and intercompartmental clearance were not estimable from
the sparse real-world data, and the interindividual variability on `ka` was so
imprecise that removing it improved the corrected Bayesian information criterion
by 7.37 units, with a further 9.48-unit improvement from fixing `ka` outright to
the value published by Gaspar 2023.

# Population

```{r population, echo = FALSE}
pop <- ui$population
tibble::tibble(
  Field = c("Species", "Subjects", "Observations", "Median age", "Age (by group)",
            "Median weight", "Female", "Renal function", "Co-medication", "Region"),
  Value = c(
    pop$species,
    paste0(pop$n_subjects, " (55 apixaban alone, 51 apixaban + amiodarone)"),
    paste0(pop$n_observations, " plasma samples, 7 below the 5 ng/mL limit of quantification"),
    pop$age_median,
    pop$age_range,
    pop$weight_median,
    paste0(pop$sex_female_pct, "%"),
    pop$renal_function,
    pop$co_medication,
    pop$regions
  )
) |>
  knitr::kable(caption = "Study population (Kolowrat 2025 Table 1 and Methods).")
```

Two features of this cohort drive everything below. It is **old** -- the group
medians are 79 and 74 years and the interquartile ranges reach 86 -- and it is
**renally impaired**, with a median estimated glomerular filtration rate of 48
mL/min/1.73 m^2 and a median Cockcroft-Gault creatinine clearance near 38 mL/min
in both arms. That is why the estimated apparent clearance of 1.5 L/h is well
below the roughly 3.3 L/h typical of younger apixaban cohorts, and why the
authors' own comparison against Morath 2025 (Figure S5) shows this model
predicting higher exposures.

The same information is available programmatically via
`readModelDb("Kolowrat_2025_apixaban")()$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Kolowrat_2025_apixaban.R`. The table below
collects them in one place for review.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Parameter, ~Value, ~Source,
  "`lka`", "log(0.82), fixed", "Table 2 final: Ka = 0.82 1/h (Fixed). Results 3.2 gives the provenance (Gaspar 2023, reference 22) and the reason (IIV on Ka imprecise; dBICc -9.48 on fixing).",
  "`lcl`", "log(1.5)", "Table 2 final: CL/F = 1.5 L/h (RSE 8.99%, 95% CI 1.26-1.78). Intercept of the Table 2 note equation at AGE = 77, no amiodarone.",
  "`lvc`", "log(45.57)", "Table 2 final: V/F = 45.57 L (RSE 10.6%, 95% CI 37.11-55.96).",
  "`e_age_cl`", "-1.52", "Table 2 final: beta_CL,age = -1.52 (RSE 25%, 95% CI -2.27 to -0.78).",
  "`e_amio_cl`", "-0.4", "Table 2 final: beta_CL,amio = -0.4 (RSE 33.8%, 95% CI -0.66 to -0.13).",
  "`etalvc`", "0.244754", "Table 2 final: IIV V = 52.66% CV (RSE 17.4%, omega CI 0.35-0.69); omega^2 = log(0.5266^2 + 1).",
  "`etalcl`", "0.327329", "Table 2 final: IIV CL = 62.23% CV (RSE 9.65%, omega CI 0.47-0.69); omega^2 = log(0.6223^2 + 1).",
  "`propSd`", "0.15", "Table 2 final: proportional error 0.15 (RSE 18.3%, 95% CI 0.11-0.21).",
  "`addSd`", "21.28", "Table 2 final: additive error 21.28 ng/mL (RSE 25.4%, 95% CI 13.21-34.26).",
  "CL/F equation", "n/a", "Table 2 note: CL/F_i = 1.5 x (Age/77)^beta_age x e^(beta_amio x 1_{CAT=1}) x e^(IIV_CL/F_i).",
  "Continuous covariate form", "n/a", "Methods 2.3: log(theta_i) = log(theta_pop) + beta_theta x log(COV_i / COV_median) + eta_theta,i.",
  "Categorical covariate form", "n/a", "Methods 2.3: log(theta_i) = log(theta_pop) + beta_theta x 1_{CAT_i=1} + eta_theta,i.",
  "One compartment, first-order absorption, no lag", "n/a", "Methods 2.2 and Results 3.2 (two-compartment models rejected: peripheral V and Q not supported).",
  "Combined proportional + additive error", "n/a", "Methods 2.2 and Results 3.2 (selected over additive-only and proportional-only)."
) |>
  knitr::kable(caption = "Source trace for every ini() value and every model equation.")
```

## How the interindividual-variability scale was pinned

Table 2 reports the interindividual variability as a percent CV (`52.66`,
`62.23`) but reports the *confidence interval for the same row* on the omega
(standard deviation) scale (`0.35-0.69`, `0.47-0.69`) -- two different scales in
one row. That makes the internal variance ambiguous on its face, so it is worth
showing how the ambiguity resolves rather than asserting it.

Monolix computes confidence intervals for variance parameters on the log scale,
`omega * exp(+/- 1.96 * RSE)`. Applying the log-normal inversion
`omega = sqrt(log(CV^2 + 1))` reproduces **all eight** printed bounds across the
base and final models; reading the CV column as omega directly misses six of
them.

```{r iiv-scale, echo = FALSE}
cv  <- c(0.5258, 0.6991, 0.5266, 0.6223)
rse <- c(0.1600, 0.0908, 0.1740, 0.0965)
lab <- c("V/F base", "CL/F base", "V/F final", "CL/F final")
om  <- sqrt(log(cv^2 + 1))
iiv <- tibble::tibble(
  Row = lab,
  `CV%` = cv * 100,
  omega = round(om, 4),
  `omega^2` = round(om^2, 6),
  Reconstructed = sprintf("%.3f - %.3f", om * exp(-1.96 * rse), om * exp(1.96 * rse)),
  Printed = c("0.36 - 0.67", "0.53 - 0.75", "0.35 - 0.69", "0.47 - 0.69"),
  Alternative = sprintf("%.3f - %.3f", cv * exp(-1.96 * rse), cv * exp(1.96 * rse))
)
iiv |>
  dplyr::rename(
    "Table 2 row"                       = Row,
    "omega = sqrt(log(CV^2+1))"         = omega,
    "variance used in ini()"            = `omega^2`,
    "CI from that omega"                = Reconstructed,
    "CI printed in Table 2"             = Printed,
    "CI if CV column were omega"        = Alternative
  ) |>
  knitr::kable(
    digits  = 2,
    caption = paste(
      "The log-normal inversion reproduces every printed Table 2 interval;",
      "treating the CV% column as omega does not."
    )
  )

# Deterministic arithmetic on printed values -- no simulation, no RNG. Every
# reconstructed bound must land on the printed bound to within its own rounding
# (Table 2 prints two decimals). Realised maximum discrepancy 0.006; the
# alternative reading below is off by up to 0.085, so this bound discriminates.
recon   <- cbind(om * exp(-1.96 * rse), om * exp(1.96 * rse))
alt     <- cbind(cv * exp(-1.96 * rse), cv * exp(1.96 * rse))
print_b <- cbind(c(0.36, 0.53, 0.35, 0.47), c(0.67, 0.75, 0.69, 0.69))
stopifnot(
  max(abs(recon - print_b)) < 0.01,
  # The gate must be able to reject: the CV-as-omega reading has to FAIL it.
  max(abs(alt - print_b)) > 0.05
)
```

# Model structure at a glance

```{r structure, echo = FALSE}
tv <- ui$theta
cl0 <- exp(tv[["lcl"]]); vc0 <- exp(tv[["lvc"]]); ka0 <- exp(tv[["lka"]])
grid <- tidyr::expand_grid(AGE = c(65, 77, 85), CONMED_AMIO = c(0L, 1L)) |>
  mutate(
    `CL/F (L/h)` = cl0 * (AGE / 77)^tv[["e_age_cl"]] * exp(tv[["e_amio_cl"]] * CONMED_AMIO),
    `V/F (L)`    = vc0,
    `kel (1/h)`  = `CL/F (L/h)` / `V/F (L)`,
    `t1/2 (h)`   = log(2) / `kel (1/h)`,
    Amiodarone   = ifelse(CONMED_AMIO == 1L, "yes", "no")
  )
grid |>
  select(AGE, Amiodarone, `CL/F (L/h)`, `V/F (L)`, `kel (1/h)`, `t1/2 (h)`) |>
  dplyr::rename("Age (years)" = AGE) |>
  knitr::kable(digits = 3, caption = "Typical-value parameters across the fitted age span.")
```

The implied terminal half-life is 21 h for a typical 77-year-old without
amiodarone and 31 h with it. That is notably longer than the roughly 12 h
apixaban half-life the paper itself invokes in Methods 2.1 when it defines a
60 h (five half-life) washout window, and it follows directly from the low
apparent clearance in this elderly, renally impaired cohort. It is a property of
the published estimates, not a transcription artefact -- see *Assumptions and
deviations*.

# Deterministic checks

These use `rxode2::zeroRe()` and steady-state dosing (`ss = 1`), so they contain
no random numbers at all and are reproducible on any machine and any thread
count. The tolerances are therefore tight: they measure numerical error, not
cohort sampling.

```{r helpers}
tau <- 12  # dosing interval (h), twice daily

# One steady-state dosing interval for a single typical subject. `ss = 1` with
# `ii = tau` asks rxode2 for the exact steady state, which avoids any question
# about whether a hand-built dosing train ran long enough (the longest half-life
# this model can produce, for an old subject on amiodarone with a low-CL eta, is
# over 100 h).
ss_events <- function(id, dose, age, amio, by = 0.05) {
  dplyr::bind_rows(
    data.frame(time = 0, amt = dose, evid = 1L, cmt = "depot",
               ss = 1L, ii = tau),
    data.frame(time = seq(0, tau, by = by), amt = NA_real_, evid = 0L,
               cmt = "central", ss = NA_integer_, ii = NA_real_)
  ) |>
    dplyr::mutate(id = id, AGE = age, CONMED_AMIO = amio) |>
    dplyr::arrange(time, dplyr::desc(evid))
}

mod_typical <- rxode2::zeroRe(ui)
```

## Mass balance: AUC over one interval equals dose divided by clearance

For any linear model at steady state, `AUC[0,tau] = Dose / CL` exactly. This is a
genuine gate on the ODE solution: the left-hand side comes from integrating the
`central` state that the solver produced, the right-hand side from the individual
clearance the covariate model computed. A wrong `kel`, a mis-wired depot, or an
absorption term that loses mass all break it.

```{r gate-massbalance}
arms <- tibble::tribble(
  ~grp,              ~dose, ~amio, ~age_t1,
  "2.5 mg BID",        2.5,    0L,      79,
  "2.5 mg BID+amio",   2.5,    1L,      74,
  "5 mg BID",          5.0,    0L,      79,
  "5 mg BID+amio",     5.0,    1L,      74
)

ev_t1 <- dplyr::bind_rows(lapply(seq_len(nrow(arms)), function(i) {
  ss_events(i, arms$dose[i], arms$age_t1[i], arms$amio[i]) |>
    dplyr::mutate(grp = arms$grp[i], dose = arms$dose[i])
}))
sim_t1 <- rxode2::rxSolve(mod_typical, events = ev_t1, keep = "grp") |>
  as.data.frame()

stopifnot(!any(is.na(sim_t1$Cc)), !any(is.nan(sim_t1$Cc)), all(sim_t1$Cc >= 0))

mb <- sim_t1 |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(grp) |>
  dplyr::summarise(
    cl_model = unique(cl),
    auc_ode  = PKNCA::pk.calc.auc(Cc, time, interval = c(0, tau), auc.type = "AUClast"),
    .groups  = "drop"
  ) |>
  dplyr::left_join(arms |> dplyr::select(grp, dose), by = "grp") |>
  dplyr::mutate(
    auc_closed = 1000 * dose / cl_model,
    pct        = 100 * (auc_ode - auc_closed) / auc_closed
  )

mb |>
  dplyr::select(grp, cl_model, auc_ode, auc_closed, pct) |>
  dplyr::rename(
    "Regimen"                 = grp,
    "CL/F (L/h)"              = cl_model,
    "AUCtau from ODE"         = auc_ode,
    "Dose / CL (closed form)" = auc_closed,
    "% difference"            = pct
  ) |>
  knitr::kable(digits = c(0, 4, 1, 1, 5),
               caption = "Steady-state mass-balance identity, ng*h/mL.")

# Numerical error only (both sides use the same drawn parameters), so a tight
# bound is correct here. Realised 0.0007% on the 0.05 h grid.
stopifnot(max(abs(mb$pct)) < 0.1)
```

## Dose linearity

The model carries no saturable term, so doubling the dose must exactly double
every exposure metric at a fixed age.

```{r gate-linearity}
ratio_noamio <- mb$auc_ode[mb$grp == "5 mg BID"]      / mb$auc_ode[mb$grp == "2.5 mg BID"]
ratio_amio   <- mb$auc_ode[mb$grp == "5 mg BID+amio"] / mb$auc_ode[mb$grp == "2.5 mg BID+amio"]
c(`no amiodarone` = ratio_noamio, `with amiodarone` = ratio_amio)

stopifnot(abs(ratio_noamio - 2) < 1e-4, abs(ratio_amio - 2) < 1e-4)
```

## The amiodarone effect reproduces the paper's headline number

The abstract, Study Highlights, Results and Discussion all state a **33%**
decrease in apixaban clearance, "ranging from 12% to 48%". That range is the
confidence interval on `beta_CL,amio` transformed the same way as the point
estimate, which is what confirms the exponential encoding
`exp(beta_amio * CONMED_AMIO)` rather than a `(1 + beta)` multiplicative
reading -- the latter would give a 40% reduction and could not produce the
12-48% range.

```{r gate-amiodarone}
beta_amio <- ui$theta[["e_amio_cl"]]
ci_amio   <- c(-0.66, -0.13)  # Table 2 final, 95% CI

amio_tab <- tibble::tibble(
  Quantity = c("Point estimate", "Lower 95% CI bound", "Upper 95% CI bound"),
  beta     = c(beta_amio, ci_amio[1], ci_amio[2]),
  Factor   = exp(beta),
  `CL reduction (%)` = 100 * (1 - Factor),
  `Paper states` = c("33%", "48%", "12%")
)
amio_tab |>
  dplyr::rename("beta_CL,amio" = beta, "exp(beta)" = Factor) |>
  knitr::kable(digits = c(0, 2, 4, 1, 0),
               caption = "Amiodarone effect on CL/F versus the paper's stated 33% (12-48%).")

# Pure arithmetic on the published coefficient; the paper prints these to whole
# percent, so agreement to 0.5 percentage points is the right tolerance.
stopifnot(
  abs(100 * (1 - exp(beta_amio))    - 33) < 0.5,
  abs(100 * (1 - exp(ci_amio[1]))   - 48) < 0.5,
  abs(100 * (1 - exp(ci_amio[2]))   - 12) < 0.5
)

# And the simulated clearances must show exactly that ratio at matched age.
cl_ratio <- mb$cl_model[mb$grp == "5 mg BID+amio"] / mb$cl_model[mb$grp == "5 mg BID"] *
  (arms$age_t1[arms$grp == "5 mg BID+amio"] / arms$age_t1[arms$grp == "5 mg BID"])^(-ui$theta[["e_age_cl"]])
stopifnot(abs(cl_ratio - exp(beta_amio)) < 1e-8)
```

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the age distribution of Table 1, which is the **only** covariate the final model
uses and the only one Table 1 stratifies. Age is drawn log-normally from each
amiodarone group's reported median and interquartile range, and the *same* age
vector is reused across the 2.5 mg and 5 mg arms within an amiodarone group so
that the dose comparison is not confounded by a different age draw.

```{r cohort}
# set.seed() seeds R's RNG (the age draw). rxSetSeed() seeds rxode2's simulation
# RNG (the etas), but rxode2 partitions its streams PER SOLVER THREAD, so the
# eta draw is reproducible on this machine and different on a machine with a
# different thread count. Every assertion below is written to hold for any
# cohort the model can produce; see pattern 12 of
# references/known-vignette-failure-patterns.md.
set.seed(20250909)
rxode2::rxSetSeed(20250909)

n_arm <- 200  # per arm; the 200/arm cap

# Log-normal age from median + IQR: sdlog = (log(Q3) - log(Q1)) / (2 * qnorm(0.75))
age_sdlog <- function(q1, q3) (log(q3) - log(q1)) / (2 * stats::qnorm(0.75))

age_groups <- tibble::tribble(
  ~amio, ~med, ~q1, ~q3,
     0L,   79,  71,  86,   # Table 1, apixaban alone
     1L,   74,  64,  83    # Table 1, apixaban + amiodarone
)
age_draw <- lapply(seq_len(nrow(age_groups)), function(i) {
  g <- age_groups[i, ]
  # Truncated to a plausible adult span; the study enrolled adults >= 18 years.
  pmin(100, pmax(18, stats::rlnorm(n_arm, log(g$med), age_sdlog(g$q1, g$q3))))
})
names(age_draw) <- as.character(age_groups$amio)

cohort_arms <- tibble::tribble(
  ~grp,              ~dose, ~amio, ~id_offset,
  "2.5 mg BID",        2.5,    0L,         0L,
  "2.5 mg BID+amio",   2.5,    1L,       200L,
  "5 mg BID",          5.0,    0L,       400L,
  "5 mg BID+amio",     5.0,    1L,       600L
)

make_cohort <- function(grp, dose, amio, id_offset) {
  subj <- tibble::tibble(
    id          = id_offset + seq_len(n_arm),
    AGE         = age_draw[[as.character(amio)]],
    CONMED_AMIO = amio,
    grp         = grp,
    dose        = dose
  )
  dplyr::bind_rows(
    subj |> dplyr::mutate(time = 0, amt = dose, evid = 1L, cmt = "depot",
                          ss = 1L, ii = tau),
    subj |> tidyr::crossing(time = seq(0, tau, by = 0.25)) |>
      dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central",
                    ss = NA_integer_, ii = NA_real_)
  ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(lapply(seq_len(nrow(cohort_arms)), function(i) {
  make_cohort(cohort_arms$grp[i], cohort_arms$dose[i],
              cohort_arms$amio[i], cohort_arms$id_offset[i])
}))
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))

# Age distributions must match Table 1 within sampling noise, and must be
# IDENTICAL between the two dose levels of the same amiodarone group.
age_chk <- events |>
  dplyr::distinct(grp, id, AGE) |>
  dplyr::group_by(grp) |>
  dplyr::summarise(median_age = median(AGE), q1 = quantile(AGE, 0.25),
                   q3 = quantile(AGE, 0.75), .groups = "drop")
age_chk |>
  dplyr::rename("Arm" = grp, "Median age" = median_age, "Q1" = q1, "Q3" = q3) |>
  knitr::kable(digits = 1, caption = "Simulated cohort age, against Table 1 (79 [71-86] and 74 [64-83]).")

stopifnot(
  identical(
    sort(age_chk$median_age[age_chk$grp == "2.5 mg BID"]),
    sort(age_chk$median_age[age_chk$grp == "5 mg BID"])
  ),
  # Median of 200 log-normal draws; realised within 1.5 years of Table 1.
  all(abs(age_chk$median_age - c(79, 74, 79, 74)[match(
    age_chk$grp, c("2.5 mg BID", "2.5 mg BID+amio", "5 mg BID", "5 mg BID+amio"))]) < 5)
)
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(ui, events = events, keep = "grp") |> as.data.frame()
stopifnot(!any(is.nan(sim$Cc)), all(sim$Cc[!is.na(sim$Cc)] >= 0))
```

# Replicate published figures

## Figure 2: simulated exposure by amiodarone receipt and dose

```{r figure-2, fig.height = 4}
# Replicates Figure 2 of Kolowrat 2025: box plots of steady-state AUCtau (A) and
# Cmin (B) stratified by receipt of amiodarone and dose.
per_id <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(grp, id) |>
  dplyr::summarise(
    auclast = PKNCA::pk.calc.auc(Cc, time, interval = c(0, tau), auc.type = "AUClast"),
    cmax    = max(Cc),
    cmin    = min(Cc),
    .groups = "drop"
  ) |>
  dplyr::mutate(grp = factor(grp, levels = cohort_arms$grp))

per_id |>
  tidyr::pivot_longer(c(auclast, cmin), names_to = "metric", values_to = "value") |>
  dplyr::mutate(metric = factor(
    metric, levels = c("auclast", "cmin"),
    labels = c("A: AUCtau (ng*h/mL)", "B: Cmin (ng/mL)")
  )) |>
  ggplot(aes(grp, value)) +
  geom_boxplot(outlier.size = 0.6, fill = "grey90") +
  facet_wrap(~metric, scales = "free_y") +
  labs(x = NULL, y = NULL,
       caption = "Replicates Figure 2 of Kolowrat 2025 (n = 200 per arm).") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

## Figure S4: dose-normalized visual predictive check

```{r figure-s4}
# Replicates Figure S4 of Kolowrat 2025: dose-normalized VPC over one steady-
# state dosing interval, stratified by receipt of amiodarone.
sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(
    dose  = ifelse(grepl("^2.5", grp), 2.5, 5),
    amio  = ifelse(grepl("amio", grp), "Apixaban + amiodarone", "Apixaban only"),
    cc_dn = Cc / dose
  ) |>
  dplyr::group_by(amio, time) |>
  dplyr::summarise(
    Q05 = quantile(cc_dn, 0.05), Q50 = quantile(cc_dn, 0.50),
    Q95 = quantile(cc_dn, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~amio) +
  labs(x = "Time after dose (h)", y = "Dose-normalized apixaban (ng/mL per mg)",
       caption = "Replicates Figure S4 of Kolowrat 2025; 5th, 50th and 95th percentiles.") +
  theme_bw()
```

The amiodarone panel sits higher and flatter, which is the visual signature of a
33% lower clearance at an unchanged volume: the trough rises more than the peak,
so the peak-to-trough ratio compresses.

# PKNCA validation

```{r pknca}
# IMPORTANT: the only filter is !is.na(Cc). Adding `time > 0` or `Cc > 0` would
# drop the interval-start row PKNCA needs to anchor AUC.
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, grp)

dose_df <- events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, grp)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | grp + id,
                             concu = "ng/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | grp + id, doseu = "mg")

# Steady-state interval: one full dosing interval starting at the dose.
intervals <- data.frame(
  start   = 0,
  end     = tau,
  cmax    = TRUE,
  tmax    = TRUE,
  cmin    = TRUE,
  auclast = TRUE,
  cav     = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_res$result |>
  as.data.frame() |>
  dplyr::group_by(grp, PPTESTCD) |>
  dplyr::summarise(median = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  dplyr::rename(
    "Arm"                = grp,
    "AUCtau (ng*h/mL)"   = auclast,
    "Cavg (ng/mL)"       = cav,
    "Cmax (ng/mL)"       = cmax,
    "Cmin (ng/mL)"       = cmin,
    "Tmax (h)"           = tmax
  ) |>
  knitr::kable(digits = 1, caption = "Cohort medians from PKNCA over one steady-state interval.")
```

# Comparison against published NCA

Kolowrat 2025 Table 3 reports median AUCtau, Cmax and Cmin for each of four
dose-by-amiodarone groups. Comparing against it requires care, because **Table 1
stratifies age by amiodarone receipt while Table 3 stratifies exposure by dose
*and* amiodarone**. The per-dose-group age distributions -- which the only
retained continuous covariate depends on -- are never reported. The two tables
below make that gap visible instead of hiding it.

## Comparison 1: typical patient at the reported group median age

This is the comparison Table 1 actually supports: a typical patient aged 79
(apixaban alone) or 74 (plus amiodarone), at each dose.

```{r nca-compare-1}
published <- tibble::tribble(
  ~grp,              ~auclast, ~cmax, ~cmin,
  "2.5 mg BID",          1823,   152,   123,
  "2.5 mg BID+amio",     2805,   231,   189,
  "5 mg BID",            2622,   266,   194,
  "5 mg BID+amio",       3626,   310,   235
)

nca_t1 <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(sim_t1 |> dplyr::filter(!is.na(Cc)) |>
                     dplyr::select(id, time, Cc, grp),
                   Cc ~ time | grp + id, concu = "ng/mL", timeu = "h"),
  PKNCA::PKNCAdose(ev_t1 |> dplyr::filter(evid == 1L) |>
                     dplyr::select(id, time, amt, grp),
                   amt ~ time | grp + id, doseu = "mg"),
  intervals = data.frame(start = 0, end = tau, cmax = TRUE, cmin = TRUE, auclast = TRUE)
))

cmp1 <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_t1,
  reference     = published,
  by            = "grp",
  units         = c(auclast = "ng*h/mL", cmax = "ng/mL", cmin = "ng/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp1, align = c("l", "l", "r", "r", "r"),
             caption = paste("Simulated (typical patient at the Table 1 group median age)",
                             "vs Kolowrat 2025 Table 3.",
                             "* differs from reference by more than 20%."))
attr(cmp1, "footnote")
```

Every 2.5 mg row agrees within 17% and most within 10%. Every 5 mg row runs
**high**, by 20% to 48%, and all three metrics in a given 5 mg arm are inflated
by a similar factor. That coherence is the diagnosis: a similar proportional
inflation of AUCtau, Cmax *and* Cmin is the signature of a single clearance
offset, not of a structural error in volume or absorption, which would move the
peak and the trough in opposite directions.

The mechanism is clinical. Apixaban is dose-reduced to 2.5 mg twice daily for
patients meeting two of three criteria, one of which is age 80 years or older,
so a 5 mg group is systematically younger than a 2.5 mg group drawn from the same
ward. Assigning both dose groups the pooled median age of their amiodarone
stratum therefore over-ages the 5 mg arms and, through an age exponent of -1.52,
understates their clearance.

## Comparison 2: age conditioned on each group's own reported AUCtau

The confound can be removed without tuning anything. At steady state
`AUCtau = Dose / CL` exactly, so each group's **own** published AUCtau pins that
group's median clearance, and inverting the age term recovers the median age the
paper did not print. Cmax and Cmin then become a genuine test of `V/F` and `ka`,
which no part of that inversion touched.

```{r implied-age}
th <- ui$theta
implied <- published |>
  dplyr::left_join(arms |> dplyr::select(grp, dose, amio), by = "grp") |>
  dplyr::mutate(
    cl_implied  = dose / (auclast / 1000),
    age_implied = 77 * (cl_implied / (exp(th[["lcl"]]) * exp(th[["e_amio_cl"]] * amio)))^(1 / th[["e_age_cl"]])
  )

implied |>
  dplyr::select(grp, auclast, cl_implied, age_implied) |>
  dplyr::rename(
    "Arm"                          = grp,
    "Published AUCtau (ng*h/mL)"   = auclast,
    "Implied median CL/F (L/h)"    = cl_implied,
    "Implied median age (years)"   = age_implied
  ) |>
  knitr::kable(digits = c(0, 0, 3, 1),
               caption = "Median age implied by each group's own published AUCtau.")

# The recovered ages are the paper's own missing stratification, and they land
# where clinical practice puts them: the dose-reduced arms about 18 years older
# than the full-dose arms, straddling the label's age-80 criterion. Both bounds
# are absolute, not taken from this run.
age_25 <- implied$age_implied[implied$dose == 2.5]
age_50 <- implied$age_implied[implied$dose == 5.0]
stopifnot(
  min(age_25) - max(age_50) > 10,        # realised 15.9 years of separation
  all(age_25 > 77), all(age_50 < 77),    # dose-reduced older, full-dose younger
  all(implied$age_implied > 40), all(implied$age_implied < 100)
)
```

The recovered medians are about 82 and 83 years for the two 2.5 mg arms and
about 66 and 63 years for the two 5 mg arms -- a 16-to-20-year split straddling
the label's age-80 dose-reduction criterion, recovered from exposure alone. That
the model's age term reproduces a clinically expected stratification the paper
never reported is itself a check on the covariate form.

```{r nca-compare-2}
ev_t2 <- dplyr::bind_rows(lapply(seq_len(nrow(implied)), function(i) {
  ss_events(i, implied$dose[i], implied$age_implied[i], implied$amio[i]) |>
    dplyr::mutate(grp = implied$grp[i])
}))
sim_t2 <- rxode2::rxSolve(mod_typical, events = ev_t2, keep = "grp") |> as.data.frame()

nca_t2 <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(sim_t2 |> dplyr::filter(!is.na(Cc)) |>
                     dplyr::select(id, time, Cc, grp),
                   Cc ~ time | grp + id, concu = "ng/mL", timeu = "h"),
  PKNCA::PKNCAdose(ev_t2 |> dplyr::filter(evid == 1L) |>
                     dplyr::select(id, time, amt, grp),
                   amt ~ time | grp + id, doseu = "mg"),
  intervals = data.frame(start = 0, end = tau, cmax = TRUE, cmin = TRUE)
))

# AUCtau is deliberately absent: it is the quantity used to pin the age, so it
# would be 0% by construction and would carry no information.
cmp2 <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_t2,
  reference     = published |> dplyr::select(grp, cmax, cmin),
  by            = "grp",
  units         = c(cmax = "ng/mL", cmin = "ng/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp2, align = c("l", "l", "r", "r", "r"),
             caption = paste("Cmax and Cmin with age conditioned on each group's own published",
                             "AUCtau, vs Kolowrat 2025 Table 3.",
                             "AUCtau is excluded because it is the conditioning quantity."))

# Deterministic (zeroRe, ss = 1), so this bound measures agreement with the
# paper's rounded published medians, not sampling noise. Realised max 12.4%.
# A 20% error in V/F or in ka moves Cmax and Cmin far outside 18%.
cmp2_pct <- as.numeric(sub("%", "", sub("\\*$", "", cmp2$`% diff`)))
stopifnot(max(abs(cmp2_pct)) < 18)
```

With age conditioned on each group's own AUCtau, **every** Cmax and Cmin row
agrees within 13% and no row is starred. `V/F = 45.57 L` and the fixed
`ka = 0.82 1/h` therefore reproduce the published peak-to-trough shape across
both dose levels and both amiodarone strata. Comparison 1's starred rows are a
reporting gap in Table 3, not a defect in the extracted model.

## Exposure ratios

```{r gmr}
matched <- exp(-ui$theta[["e_amio_cl"]])  # amiodarone effect at matched age
ratios <- tibble::tibble(
  Metric = c("AUCtau", "Cmax", "Cmin"),
  `Paper GMR, 2.5 mg` = c(1.58, 1.48, 1.68),
  `Paper GMR, 5 mg`   = c(1.12, 1.09, 1.11),
  `Model, matched age` = c(matched, NA, NA)
)
ratios |>
  knitr::kable(digits = 3, caption = paste(
    "Published geometric mean ratios (amiodarone vs none) against the model's",
    "matched-age AUC ratio exp(0.4)."))
```

The model's matched-age AUCtau ratio is `exp(0.4) = 1.492`, which sits between
the paper's two reported AUCtau geometric mean ratios (1.58 at 2.5 mg, 1.12 at
5 mg). Those two GMRs are not estimates of the same quantity as the model
coefficient: each compares two groups that also differ in age, and the paper's
own confidence intervals -- [1.28, 1.87] and [0.91, 1.34] -- overlap 1.492 in
the 2.5 mg case and come close in the 5 mg case. The authors make the same point
in the Discussion, noting that "the extent of the apixaban elevations was not
necessarily maintained between dose levels". No gate is placed on these ratios
because both sides are confounded by the unreported per-dose-group age
distribution.

## Cohort comparison

For completeness, the same Table 3 comparison against the stochastic cohort
medians rather than a typical patient.

```{r nca-compare-cohort}
cmp3 <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "grp",
  units         = c(auclast = "ng*h/mL", cmax = "ng/mL", cmin = "ng/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp3, align = c("l", "l", "r", "r", "r"),
             caption = paste("Cohort medians (n = 200 per arm, age from Table 1)",
                             "vs Kolowrat 2025 Table 3.",
                             "* differs from reference by more than 20%."))

# Cohort medians, so the bound must admit the eta draw as well as the age draw.
# The 2.5 mg arms are the ones whose age distribution Table 1 actually supports;
# the 5 mg arms are a known, explained deviation and are excluded from the gate
# (see Comparison 1). Realised max 17% on the 2.5 mg arms.
cmp3_pct <- as.numeric(sub("%", "", sub("\\*$", "", cmp3$`% diff`)))
keep_25 <- grepl("^2.5", cmp3$grp)
stopifnot(max(abs(cmp3_pct[keep_25])) < 35)
```

# Assumptions and deviations

## Reporting inconsistencies in the source (errata)

- **Table 2 mixes scales within a single row.** The `IIV V (CV%)` and
  `IIV CL (CV%)` rows print the estimate as a percent CV (52.66, 62.23) but the
  adjacent `95% CI` cells on the omega (standard deviation) scale (0.35-0.69,
  0.47-0.69). The `Proportional error (CV%)` row is headed the same way but its
  value (0.15) and interval (0.11-0.21) are **fractions**, not percents. Both
  readings are recoverable and are pinned in the *Source trace* section above by
  reconstructing all eight printed variance intervals; a 0.15% proportional
  error alongside a 21.28 ng/mL additive term is not physically plausible in any
  case.
- **The additive-error row carries no unit.** It is in the units of the
  observation, ng/mL, which the 5 ng/mL assay lower limit of quantification and
  the ng/mL Cmax and Cmin of Table 3 confirm.
- **Table 1's race counts do not sum to the group N in the amiodarone column.**
  The listed counts are 29 + 12 + 2 + 1 + 1 = 45 against a stated n = 51, and
  the Asian row prints "1 (4.1)" where 1 of 51 is 2.0%. The apixaban-alone
  column sums correctly (33 + 18 + 1 + 3 + 0 = 55, percentages to 100). The
  `race_ethnicity` entry in the model's `population` metadata uses the pooled
  counts over the stated denominator of 106; race is not a covariate in the
  model, so nothing downstream depends on this.
- **The overall median age of 77 is not in Table 1.** It appears only in the
  Results 3.2 text and in the Table 2 note, while Table 1 reports the two
  amiodarone groups separately (79 and 74). The model centres on 77, per the
  paper's own equation.
- **The fitted age range is not reported**, only interquartile ranges spanning
  64-86 years. The age term is a power function with a steep exponent of -1.52,
  so extrapolating below the fitted range inflates clearance sharply (at age 40
  the multiplier is about 2.6-fold its value at 77). Treat the model as
  applicable to the elderly cohort it was built from.

## Encoding decisions

- **Combined residual error, `combined1` versus `combined2`.** The source used
  Monolix, which offers two combined parameterisations -- `combined1`
  (SD = `a + b*f`) and `combined2` (SD = `sqrt(a^2 + (b*f)^2)`) -- and the paper
  names only "combined proportional and additive error" (Methods 2.2) without
  saying which. The model file uses nlmixr2's default `add() + prop()`, which is
  the `combined2` form, because the source makes no explicit `combined1`
  declaration. This affects only the residual-error magnitude at a given
  predicted concentration, not the structural model or any deterministic check
  in this vignette; a user refitting to their own data should confirm which form
  they want.
- **No IIV correlation.** The paper reports no covariance or correlation between
  the CL/F and V/F random effects, so the two etas are entered as independent
  diagonal elements rather than a block.
- **No bioavailability term.** `F` is not identifiable from oral-only data and is
  absorbed into the apparent parameters `CL/F` and `V/F`, so no `f(depot)` is
  applied. The paper reports apparent parameters throughout.
- **`ka` is fixed, and its provenance is a second paper.** `ka = 0.82 1/h` was
  not estimated here; Results 3.2 states it was fixed to the value from
  reference 22, which is Gaspar 2023. That value is independently confirmed
  against the sibling extraction `modellib("Gaspar_2023_apixaban")`, whose own
  Table 2 reports `ka = 0.82 1/h` (RSE 11%). The `reference` field of this model
  file cites both papers.
- **Base model not extracted.** Table 2 also reports a covariate-free base model
  (V/F 43.45 L, CL/F 1.32 L/h). Per the library's replicate-the-author's-
  structure policy only the final model is packaged.
- **The implied half-life exceeds the value the paper assumes elsewhere.** This
  model gives a terminal half-life of 21 h at age 77 without amiodarone and 31 h
  with it, whereas Methods 2.1 invokes a 60 h washout as "five half-lives",
  i.e. about 12 h. The 12 h figure is the label value cited from reference 1, not
  an output of this model; the longer half-life follows arithmetically from the
  low apparent clearance (1.5 L/h) fitted in this elderly, renally impaired
  cohort. No change was made to the extracted parameters.

## Simulation assumptions

- **Age distribution.** Drawn log-normally from each amiodarone group's Table 1
  median and interquartile range, truncated to 18-100 years. Table 1 reports no
  full range and no distributional form.
- **Per-dose-group age is unknown.** Table 3 stratifies exposure by dose and
  amiodarone, but Table 1 stratifies age only by amiodarone. Comparison 1 assigns
  each dose group its amiodarone stratum's pooled median age, which is why its
  5 mg rows are starred; Comparison 2 recovers the missing stratification from the
  paper's own AUCtau. Neither comparison changes any model parameter.
- **Covariates held constant.** The source used only the first reported value per
  encounter (Limitations), so age is time-fixed here, matching the fitted model.
- **Steady state via `ss = 1`.** All patients were assumed to be at steady state
  with apixaban and amiodarone (Methods 2.1). `ss = 1` with `ii = 12` gives the
  exact steady state and avoids any question about whether a hand-built dosing
  train ran long enough -- relevant here because an old subject on amiodarone
  with a low-clearance eta can have a half-life over 100 h.
- **Cohort size** is 200 per arm, the library cap; the paper simulated 104 actual
  and 1000 sampled individuals.
- **BLQ handling.** Seven of 360 observations were below the 5 ng/mL limit of
  quantification and were censored during estimation. The model emits continuous
  concentrations, so no BLQ rule is applied in simulation; predicted troughs in
  this cohort are far above 5 ng/mL in any case.
