---
title: "Digoxin (Komatsu 2015)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Digoxin (Komatsu 2015)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r readModelDb("Komatsu_2015_digoxin")()$reference`
* Description: `r readModelDb("Komatsu_2015_digoxin")()$description`
* Article: <https://doi.org/10.1186/s40780-015-0023-6>

```{r load}
mod <- readModelDb("Komatsu_2015_digoxin")
tv  <- rxode2::zeroRe(readModelDb("Komatsu_2015_digoxin"))
```

# Population

Komatsu 2015 analysed 287 steady-state trough serum digoxin concentrations from
192 adult Japanese cardiology inpatients and outpatients treated at Kitasato
University Hospital between November 2011 and January 2013 (Table 1). The cohort
was elderly (mean age 71 +/- 12 years), light (mean weight 55.47 +/- 11.94 kg)
and renally impaired relative to a healthy adult population (mean
Cockcroft-Gault creatinine clearance 56.17 +/- 33.76 mL/min). 121 of 192 were
male (37.0% female). Left-ventricular ejection fraction was at or above 40% in
156 patients and below 40% in 36. Patients with major hepatic or
gastrointestinal disorders, on dialysis, or with rapidly deteriorating renal
function were excluded.

Maintenance dosing ranged from 0.125 mg every 3 days to 0.25 mg/day, with
0.125 mg/day accounting for 200 of the 287 observations. 15 patients received
concomitant amiodarone. Observed trough concentrations averaged
0.90 +/- 0.56 ng/mL. Samples were drawn before the morning dose at least one
week after digoxin was started, and were assayed by cloned enzyme immunoassay
with a 0.2 ng/mL minimum detectable concentration and intra- and inter-assay
CV below 10%.

The same information is available programmatically via
`readModelDb("Komatsu_2015_digoxin")()$population`.

# Structural model: why there is no ODE and no NCA

This is an unusually minimal population PK model, and the reason is in the data.
Komatsu 2015 used routine therapeutic-drug-monitoring troughs drawn only after
steady state had been reached. There is no absorption phase and no distribution
phase anywhere in the dataset, so the authors fit neither. Their Methods section
states the structural model in full as the algebraic steady-state identity

$$C_{ss,ij} = \frac{D_{ij}}{CL_{ij}\,\tau_{ij}}$$

and their Discussion says so explicitly: *"our model can't individually assess
these parameters because our data is trough concentration sampling"* and *"our
population model didn't consider volume of distribution or absorption phase."*

Two consequences follow, and both are deliberate:

1. **The packaged model has no `d/dt()` state, no volume and no `ka`.**
   Inventing a volume of distribution to make it look like a conventional
   one-compartment model would be fabricating a parameter the paper never
   estimated. `modeldb` records this model with `algebraic = TRUE`.
2. **PKNCA is not used in this vignette.** Non-compartmental analysis operates
   on a concentration-time profile, and this model produces a single
   steady-state trough per subject rather than a profile. Running PKNCA here
   would be a gate that cannot go red. The validation below substitutes checks
   that do have the power to fail: an exact mass-balance identity, a
   reproduction of the paper's own published dosing nomogram from its own
   parameters, and a cohort-level comparison against the observed
   concentrations in Table 1.

The dose enters as the data column `DOSE_DIGOXIN_MGD` (mg/day) rather than as an
`amt` on an event record, because there is no compartment to dose into. Any
digoxin model in this package that *does* carry disposition structure
(`Zhou_2010_digoxin`, `Jelliffe_2014_digoxin`, `Gu_2025_digoxin_pbpk`) doses
through the event table instead.

# Source trace

Every `ini()` entry carries an in-file comment pointing at its source location
in `inst/modeldb/specificDrugs/Komatsu_2015_digoxin.R`. They are collected here
for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `Cc <- 1000 * DOSE_DIGOXIN_MGD / (24 * cl)` | n/a | Methods, *Pharmacokinetics model*: `Cssij = Dij/(CLij tau_ij)`. The 24 converts CL from L/h to L/day; the 1000 converts mg/L to ng/mL. |
| `cl <- (exp(lcl) + e_crcl_cl * CRCL) * (1 + e_amio_cl * amd) * exp(etalcl)` | n/a | Results final model and Table 3: `CL/F (L/h) = (1.21 + 0.0532 x CLcr [mL/min]) x (1 + 0.787 x AMD)`. Exponential IIV per Methods: `Pi = TV(Pi) x exp(eta_i)`. |
| `amd <- 1 - CONMED_AMIO` | n/a | Abstract / Results: *"AMD is 0 in the case of concomitant administration of amiodarone and 1 otherwise"* -- the inverse of the canonical `CONMED_AMIO` orientation. |
| `lcl` (`exp(lcl)` = 1.21 L/h) | `log(1.21)` | Table 4, theta 1 = 1.21 +/- 0.21 (bootstrap 1.30, 95% CI 1.20-1.39); same value in the Results equation and Table 3. |
| `e_crcl_cl` | 0.0532 | Table 4, theta 2 = 0.0532 +/- 0.0068 (bootstrap 0.0543, 95% CI 0.050-0.057). |
| `e_amio_cl` | 0.787 | Table 4, theta 3 = 0.787 +/- 0.187 (bootstrap 0.803, 95% CI 0.768-0.838). |
| `etalcl` (variance) | 0.104 | Table 4, `omega CL` final-model column = 0.104 +/- 0.017. Cross-checks against Table 3 / Results `omega(CL) = 32.2%`, since `sqrt(0.104) = 0.3225`. |
| `propSd` | `sqrt(0.065)` = 0.2550 | Table 4, `sigma` final-model column = 0.065 +/- 0.010. Cross-checks against Table 3 / Results `sigma = 25.5%`, since `sqrt(0.065) = 0.2550`. Proportional form per Methods: `Cobs = Cpred x (1 + eps)`. |

Note on Table 4: the **final-model** column reports `omega CL` and `sigma` on the
**variance** scale (0.104, 0.065) while the **bootstrap** column reports the same
two quantities on the **standard-deviation** scale (0.324, 0.340). Reading the
bootstrap column as a second variance estimate would inflate both by roughly a
factor of three. The Table 3 / Results percentages (32.2%, 25.5%) are the
square roots of the final-model variances and settle which scale is which.

# Validation

## 1. The steady-state identity holds exactly

The model is an algebraic rearrangement of `Dose = CL x Css x tau`, so the
identity must close to machine precision for every subject. This gate catches a
unit-conversion error in either constant.

```{r massbalance}
grid_mb <- expand.grid(
  CRCL             = c(5, 20, 56.17, 90, 130),
  CONMED_AMIO      = c(0, 1),
  DOSE_DIGOXIN_MGD = c(0.03125, 0.0625, 0.125, 0.1875, 0.25)
) |>
  mutate(id = row_number(), time = 0, evid = 0)

mb <- rxode2::rxSolve(
  tv, grid_mb,
  keep = c("CRCL", "CONMED_AMIO", "DOSE_DIGOXIN_MGD"), returnType = "data.frame"
) |>
  mutate(dose_recovered = Cc * cl * 24 / 1000,
         resid = dose_recovered - DOSE_DIGOXIN_MGD)

stopifnot(
  nrow(mb) == nrow(grid_mb),      # guard: the gate had rows to test
  nrow(mb) == 50L,
  all(is.finite(mb$Cc)),
  max(abs(mb$resid)) < 1e-10
)
cat("max |Dose recovered - Dose administered| =", format(max(abs(mb$resid))), "mg/day\n")
```

## 2. Clearance as a function of creatinine clearance

Reproduces the Results final-model equation directly. The amiodarone-free line
is `1.787 x (1.21 + 0.0532 x CLcr)`; the amiodarone line is
`1.21 + 0.0532 x CLcr`.

```{r clcurve}
cl_curve <- expand.grid(CRCL = seq(5, 130, by = 1), CONMED_AMIO = c(0, 1)) |>
  mutate(id = row_number(), time = 0, evid = 0, DOSE_DIGOXIN_MGD = 0.125)

cl_out <- rxode2::rxSolve(
  tv, cl_curve, keep = c("CRCL", "CONMED_AMIO"), returnType = "data.frame"
) |>
  mutate(Amiodarone = ifelse(CONMED_AMIO == 1, "Coadministered", "Absent"))

# Closed-form cross-check of every point on both lines.
cl_hand <- with(cl_out, (1.21 + 0.0532 * CRCL) * (1 + 0.787 * (1 - CONMED_AMIO)))
stopifnot(nrow(cl_out) == 252L, max(abs(cl_out$cl - cl_hand)) < 1e-10)

ggplot(cl_out, aes(CRCL, cl, colour = Amiodarone)) +
  geom_line(linewidth = 1) +
  labs(x = "Creatinine clearance (mL/min)", y = "CL/F (L/h)",
       title = "Digoxin apparent oral clearance (Komatsu 2015 final model)") +
  theme_bw()
```

## 3. The amiodarone effect is a 1.787-fold rise in trough

The paper states in the Discussion that *"amiodarone increased the trough level
of digoxin concentration by approximately two-fold"*. Because the amiodarone
factor is purely multiplicative on clearance, the trough ratio is exactly
`1 + 0.787` at every creatinine clearance and every dose.

```{r amioeffect}
amio_ratio <- mb |>
  select(CRCL, CONMED_AMIO, DOSE_DIGOXIN_MGD, Cc) |>
  pivot_wider(names_from = CONMED_AMIO, values_from = Cc,
              names_prefix = "amio_") |>
  mutate(ratio = amio_1 / amio_0)

stopifnot(
  nrow(amio_ratio) == 25L,                       # guard: 5 CRCL x 5 doses
  max(abs(amio_ratio$ratio - 1.787)) < 1e-10
)
cat("Trough ratio (amiodarone / no amiodarone):",
    format(unique(round(amio_ratio$ratio, 6))), "\n")
```

## 4. The published dosing recommendations land inside the target range

This is the strongest available check, because the Abstract's five dosing
recommendations are a *downstream* product of the model that sits nowhere near
the parameter table. If the three thetas, the amiodarone polarity, or the
`1000 / 24` unit conversion were wrong, the recommended doses would not put a
typical patient inside the paper's own 0.5-0.8 ng/mL target band.

The `AMD` labels attached to the Abstract's dosing list are the **inverse** of
the `AMD` convention the Abstract defines two sentences earlier for the model
equation; see *Assumptions and deviations* below. The rows are labelled here by
the amiodarone status that the arithmetic actually supports.

```{r nomogram}
rec <- tibble::tribble(
  ~regimen,                 ~amio, ~crcl_lo, ~crcl_hi,    ~mgd,
  "No amiodarone, CLcr < 35",    0,        5,       35, 0.0625,
  "No amiodarone, CLcr 35-65",   0,       35,       65, 0.1250,
  "No amiodarone, CLcr 65-100",  0,       65,      100, 0.1875,
  "Amiodarone, CLcr < 30",       1,        5,       30, 0.03125,
  "Amiodarone, CLcr 30-85",      1,       30,       85, 0.0625
) |>
  mutate(CRCL = (crcl_lo + crcl_hi) / 2, id = row_number(),
         time = 0, evid = 0,
         CONMED_AMIO = amio, DOSE_DIGOXIN_MGD = mgd)

nom <- rxode2::rxSolve(
  tv, rec, keep = c("CRCL", "CONMED_AMIO", "DOSE_DIGOXIN_MGD"),
  returnType = "data.frame"
) |>
  left_join(rec |> select(id, regimen), by = "id") |>
  mutate(`In 0.5-0.8 ng/mL` = Cc >= 0.5 & Cc <= 0.8)

stopifnot(
  nrow(nom) == 5L,                          # guard: all five rows evaluated
  all(rec$regimen %in% nom$regimen),
  all(nom$`In 0.5-0.8 ng/mL`)
)

nom |>
  select(regimen, CRCL, DOSE_DIGOXIN_MGD, cl, Cc, `In 0.5-0.8 ng/mL`) |>
  rename("Published regimen" = regimen,
         "CLcr at band midpoint (mL/min)" = CRCL,
         "Dose (mg/day)" = DOSE_DIGOXIN_MGD,
         "CL/F (L/h)" = cl,
         "Typical trough (ng/mL)" = Cc) |>
  knitr::kable(digits = c(0, 1, 5, 2, 3, 0),
               caption = "Komatsu 2015 Abstract / Fig. 4 dosing recommendations, evaluated at the midpoint of each creatinine-clearance band with the packaged model's typical-value parameters.")
```

All five recommendations put the typical patient between 0.61 and 0.79 ng/mL,
inside the paper's 0.5-0.8 ng/mL target.

## 5. Probability of a trough inside the target band (replicates Figure 3)

Figure 3 of Komatsu 2015 plots the probability that a steady-state trough falls
within 0.5-0.8 ng/mL as a function of creatinine clearance, with and without
amiodarone. Because the only between-subject variability in this model is a
single log-normal `eta` on clearance, that probability has a closed form:
the individual trough is `Css_tv * exp(-eta)`, which is log-normal with
log-median `log(Css_tv)` and log-scale SD `omega = sqrt(0.104)`. The simulated
cohort below is checked against that closed form, which validates the
simulation setup rather than merely re-plotting it.

```{r fig3-analytic}
omega <- sqrt(0.104)
p_in_band <- function(css_tv, lo = 0.5, hi = 0.8) {
  pnorm((log(hi) - log(css_tv)) / omega) - pnorm((log(lo) - log(css_tv)) / omega)
}

curve_grid <- expand.grid(
  CRCL             = seq(5, 130, by = 1),
  CONMED_AMIO      = c(0, 1),
  DOSE_DIGOXIN_MGD = c(0.0625, 0.125, 0.1875, 0.25)
) |>
  mutate(id = row_number(), time = 0, evid = 0)

curve <- rxode2::rxSolve(
  tv, curve_grid,
  keep = c("CRCL", "CONMED_AMIO", "DOSE_DIGOXIN_MGD"), returnType = "data.frame"
) |>
  mutate(p = p_in_band(Cc),
         Amiodarone = ifelse(CONMED_AMIO == 1, "Coadministered", "Absent"),
         Dose = factor(paste0(DOSE_DIGOXIN_MGD, " mg/day")))

stopifnot(nrow(curve) == nrow(curve_grid), all(is.finite(curve$p)))

ggplot(curve, aes(CRCL, 100 * p, colour = Dose)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~Amiodarone) +
  labs(x = "Creatinine clearance (mL/min)",
       y = "Probability trough in 0.5-0.8 ng/mL (%)",
       title = "Replicates Figure 3 of Komatsu 2015") +
  theme_bw()
```

Now the same probabilities from a simulated cohort. Eight arms of 200 subjects
each (within the 200-per-arm cap) are drawn with between-subject variability on
clearance; the realised fraction inside the band is compared against the
analytic value. The tolerance is set from the binomial standard error at
`n = 200` (about 3.5 percentage points) with generous headroom, so the gate is
insensitive to which cohort `rxSolve` happens to draw on a given machine while
still failing on a structural error.

```{r fig3-simulated}
rxode2::rxSetSeed(20150713)
set.seed(20150713)

n_per_arm <- 200L
arms <- expand.grid(CRCL = c(20, 50, 80, 110), CONMED_AMIO = c(0, 1))
arms$arm <- paste0("CLcr ", arms$CRCL,
                   ifelse(arms$CONMED_AMIO == 1, " + amiodarone", ""))

sim_dat <- arms |>
  rowwise() |>
  do(tibble(
    arm = .$arm, CRCL = .$CRCL, CONMED_AMIO = .$CONMED_AMIO,
    id = seq_len(n_per_arm)
  )) |>
  ungroup() |>
  mutate(id = row_number(), time = 0, evid = 0, DOSE_DIGOXIN_MGD = 0.125)

sim_out <- rxode2::rxSolve(
  mod, sim_dat, keep = c("arm", "CRCL", "CONMED_AMIO", "DOSE_DIGOXIN_MGD"),
  returnType = "data.frame"
)

sim_p <- sim_out |>
  group_by(arm, CRCL, CONMED_AMIO) |>
  summarise(n = n(), p_sim = mean(Cc >= 0.5 & Cc <= 0.8), .groups = "drop") |>
  mutate(css_tv = 1000 * 0.125 /
           (24 * (1.21 + 0.0532 * CRCL) * (1 + 0.787 * (1 - CONMED_AMIO))),
         p_analytic = p_in_band(css_tv),
         abs_diff_pct = 100 * abs(p_sim - p_analytic))

stopifnot(
  nrow(sim_p) == 8L,                    # guard: every arm produced a row
  all(sim_p$n == n_per_arm),
  max(sim_p$abs_diff_pct) < 12
)

sim_p |>
  select(arm, p_analytic, p_sim, abs_diff_pct) |>
  mutate(across(c(p_analytic, p_sim), ~ 100 * .x)) |>
  rename("Arm" = arm, "Analytic (%)" = p_analytic,
         "Simulated, n = 200 (%)" = p_sim,
         "Absolute difference (pp)" = abs_diff_pct) |>
  knitr::kable(digits = 1,
               caption = "Simulated cohort probabilities against the closed-form log-normal result, 0.125 mg/day.")
```

## 6. Reconstructing the Figure 4 nomogram from the model

Figure 4 of Komatsu 2015 is a nomogram: for each creatinine clearance, the dose
that best hits the 0.5-0.8 ng/mL band. Selecting, at each creatinine clearance,
whichever available dose strength maximises the closed-form probability from
section 5 reconstructs that nomogram from the model alone.

```{r fig4}
dose_grid <- c(0.03125, 0.0625, 0.125, 0.1875, 0.25)
best_dose <- function(crcl, amio) {
  css <- 1000 * dose_grid /
    (24 * (1.21 + 0.0532 * crcl) * (1 + 0.787 * (1 - amio)))
  dose_grid[which.max(p_in_band(css))]
}

nomo <- expand.grid(CRCL = seq(5, 130, by = 5), CONMED_AMIO = c(0, 1)) |>
  rowwise() |>
  mutate(best = best_dose(CRCL, CONMED_AMIO)) |>
  ungroup()

# The reconstructed band edges. `edge()` fails loudly if a transition is absent,
# so a band that never appears cannot be silently scored as agreeing.
edge <- function(amio, dose) {
  v <- nomo$CRCL[nomo$CONMED_AMIO == amio & nomo$best == dose]
  if (!length(v)) stop("dose ", dose, " is never selected at amio = ", amio)
  c(lo = min(v), hi = max(v))
}

bands <- tibble::tibble(
  Arm = c("No amiodarone", "No amiodarone", "No amiodarone",
          "Amiodarone", "Amiodarone"),
  `Dose (mg/day)` = c(0.0625, 0.125, 0.1875, 0.03125, 0.0625),
  `Published band (mL/min)` = c("< 35", "35-65", "65-100", "< 30", "30-85"),
  `Reconstructed band (mL/min)` = c(
    paste(edge(0, 0.0625),  collapse = "-"),
    paste(edge(0, 0.125),   collapse = "-"),
    paste(edge(0, 0.1875),  collapse = "-"),
    paste(edge(1, 0.03125), collapse = "-"),
    paste(edge(1, 0.0625),  collapse = "-")
  )
)

# The two amiodarone bands reproduce both published edges exactly (30 and 85),
# and the no-amiodarone 0.0625 band reproduces its published upper edge (35).
stopifnot(
  nrow(bands) == 5L,
  edge(1, 0.03125)[["hi"]] == 30,
  edge(1, 0.0625)[["hi"]]  == 85,
  edge(0, 0.0625)[["hi"]]  == 35,
  # The dose LEVELS are exactly the published set in both arms.
  setequal(nomo$best[nomo$CONMED_AMIO == 1], c(0.03125, 0.0625, 0.125)),
  all(c(0.0625, 0.125, 0.1875) %in% nomo$best[nomo$CONMED_AMIO == 0])
)

knitr::kable(bands,
  caption = "Published dosing bands against bands reconstructed by maximising the closed-form in-band probability over the available dose strengths.")
```

Three of the five published band edges are reproduced exactly. The
0.125 -> 0.1875 mg/day transition in the amiodarone-free arm is the exception
and is discussed under *Assumptions and deviations*.

## 7. Virtual cohort against the observed concentrations in Table 1

A 192-subject virtual cohort is drawn to match the Table 1 marginals: creatinine
clearance from `N(56.17, 33.76)` truncated to the 5-130 mL/min range over which
the paper ran its own simulations, amiodarone in 15/192 subjects, and daily dose
sampled from the Table 1 regimen counts. Simulated concentrations carry both
between-subject variability and the proportional residual error, and are
censored at half the assay's 0.2 ng/mL minimum detectable concentration, so they
are comparable with the observed `0.90 +/- 0.56 ng/mL`.

```{r cohort}
rxode2::rxSetSeed(20150713)
set.seed(20150713)

n_sub <- 192L
# Table 1 regimen counts, expressed as daily dose. 0.125 mg every 2 days and
# 0.0625 mg/day are the same daily dose and are pooled (17 + 14 = 31).
regimens <- c(0.04167, 0.0625, 0.125, 0.25)
reg_wt   <- c(7,        31,     200,   49)

cohort <- tibble::tibble(
  id          = seq_len(n_sub),
  CRCL        = pmin(pmax(rnorm(n_sub, 56.17, 33.76), 5), 130),
  CONMED_AMIO = as.numeric(seq_len(n_sub) %in% sample.int(n_sub, 15L)),
  DOSE_DIGOXIN_MGD = sample(regimens, n_sub, replace = TRUE, prob = reg_wt),
  time = 0, evid = 0
)

coh <- rxode2::rxSolve(
  mod, cohort, keep = c("CRCL", "CONMED_AMIO", "DOSE_DIGOXIN_MGD"),
  returnType = "data.frame"
) |>
  mutate(conc = pmax(sim, 0.2 / 2))

stopifnot(
  nrow(coh) == n_sub,                  # guard: every subject solved
  !anyNA(coh$conc),
  all(coh$conc > 0)
)

summary_tbl <- tibble::tibble(
  Statistic = c("Mean (ng/mL)", "SD (ng/mL)", "Median (ng/mL)",
                "10th percentile", "90th percentile"),
  Simulated = c(mean(coh$conc), sd(coh$conc), median(coh$conc),
                quantile(coh$conc, 0.10), quantile(coh$conc, 0.90)),
  `Komatsu 2015 Table 1` = c(0.90, 0.56, NA, NA, NA)
)
knitr::kable(summary_tbl, digits = 3,
  caption = "Virtual-cohort steady-state troughs against the observed distribution in Table 1.")

# Centre-based assertions only. The mean of a random 192-subject cohort is not
# reproducible across rxode2 builds and thread counts, so the bound is on the
# CENTRE against the paper's own reported mean, with headroom that a
# mis-transcribed theta, dose or unit conversion (which move the whole
# distribution by tens of percent) would still blow.
stopifnot(
  abs(mean(coh$conc) - 0.90) < 0.30,
  median(coh$conc) > 0.3, median(coh$conc) < 1.4
)

ggplot(coh, aes(conc)) +
  geom_histogram(bins = 30, fill = "grey70", colour = "white") +
  geom_vline(xintercept = 0.90, colour = "firebrick", linewidth = 1) +
  annotate("rect", xmin = 0.5, xmax = 0.8, ymin = -Inf, ymax = Inf,
           alpha = 0.15, fill = "steelblue") +
  labs(x = "Simulated steady-state trough (ng/mL)", y = "Subjects",
       title = "Virtual cohort (n = 192) matching Komatsu 2015 Table 1",
       subtitle = paste("Red line: observed mean 0.90 ng/mL.",
                        "Blue band: 0.5-0.8 ng/mL target.")) +
  theme_bw()
```

The simulated mean sits close to the observed 0.90 ng/mL, and the spread is
comparable, which is the most that can be asked of a cohort reconstructed from
published marginals rather than the original covariate joint distribution.

# Assumptions and deviations

* **The amiodarone indicator is inverted relative to the canonical column.**
  Komatsu 2015 defines its own `AMD` as *0 when amiodarone is coadministered
  and 1 otherwise*, the opposite polarity to the canonical `CONMED_AMIO`
  (1 = coadministered). The model reconstructs the paper's indicator as
  `amd <- 1 - CONMED_AMIO` so that the published coefficients 1.21, 0.0532 and
  0.787 are used verbatim with no re-parameterisation. The polarity is not an
  assumption: the paper's own statement that amiodarone raises the trough
  roughly two-fold, the mechanism it cites (inhibition of renal tubular
  secretion and of P-glycoprotein), and the arithmetic in section 4 all agree
  that the factor `1 + 0.787` applies when amiodarone is **absent**.

* **The `AMD` labels on the Abstract's dosing recommendations are swapped
  relative to the `AMD` convention the Abstract itself defines.** The Abstract
  defines `AMD = 0` as amiodarone-present for the model equation, then lists
  *"0.125 mg/day (CLcr, 35-65 mL/min and AMD = 0)"* and *"0.0625 mg/day
  (CLcr, 30-85 mL/min and AMD = 1)"* among its recommendations. Taken at the
  Abstract's own definition, the first would give a typical trough of
  1.35 ng/mL and the second 0.38 ng/mL, both well outside the paper's
  0.5-0.8 ng/mL target. Read with the labels swapped -- higher doses for
  patients *not* on amiodarone, as the pharmacology requires -- all five
  recommendations land between 0.61 and 0.79 ng/mL (section 4). This vignette
  and the model file therefore follow the model equation's stated convention,
  and label the dosing recommendations by the amiodarone status the arithmetic
  supports. No parameter value is affected by this reading.

* **The 0.1875 mg/day band edge is not reproducible from the published model.**
  The Abstract places the amiodarone-free 0.1875 mg/day band at CLcr 65-100
  mL/min. Maximising the in-band probability over the available dose strengths
  puts the 0.125 -> 0.1875 transition at 80-85 mL/min instead (section 6), and
  the result is unchanged whether between-subject variability alone or
  between-subject plus residual variability is used, and unchanged if the
  criterion is switched to "typical trough closest to 0.7 ng/mL" (the target
  the Discussion cites when comparing against the Koup and Jusko methods).
  At CLcr = 65 with no amiodarone, 0.125 mg/day gives a typical trough of
  0.62 ng/mL, inside the band, while 0.1875 mg/day gives 0.94 ng/mL, outside
  it. The published band edge appears to have been read off Figure 3 by eye
  rather than computed. No parameter is adjusted to close this gap. Relatedly,
  the Abstract's recommendation list stops at CLcr 100 mL/min in the
  amiodarone-free arm even though its simulations ran to 130 mL/min.

* **No volume of distribution, absorption rate or peripheral compartment.**
  These are absent from the model because the paper did not estimate them, not
  because they were omitted in transcription; see *Structural model* above. The
  model predicts steady-state troughs only and must not be used to simulate a
  concentration-time profile, a loading dose, or a non-steady-state regimen.

* **No PKNCA validation.** Non-compartmental analysis requires a
  concentration-time profile, which this model cannot produce. The substituted
  checks are described under *Structural model*.

* **Creatinine clearance is raw Cockcroft-Gault mL/min, not BSA-normalized.**
  The canonical `CRCL` column accommodates both conventions; the per-model
  `covariateData[[CRCL]]$units` records `mL/min`. Because the covariate term is
  additive and uncentred, `exp(lcl)` = 1.21 L/h is the clearance extrapolated to
  CLcr = 0 rather than a cohort typical value, and the model should be exercised
  over the 5-130 mL/min range the paper simulated. Patients on dialysis were
  excluded, so the model is not informed at the extreme low end.

* **The virtual cohort is a reconstruction from published marginals.**
  Komatsu 2015 reports mean and SD for creatinine clearance and counts for the
  dosing regimens and amiodarone use, but not the joint distribution, and the
  original data are not public. Creatinine clearance is drawn as a truncated
  normal and the covariates are drawn independently, which will understate any
  real correlation between renal function and prescribed dose.

* **Body weight, age and ejection fraction are not in the model.** Body weight
  was significant in univariate screening (Table 2, LLD 10.234, p < 0.002) but
  did not survive backward elimination; age and ejection fraction were not
  significant at any stage. They are recorded in `population` but are not
  `covariateData` entries, since the final model does not reference them.
