---
title: "Docetaxel + ritonavir (Koolen 2010)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Docetaxel + ritonavir (Koolen 2010)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r readModelDb("Koolen_2010_docetaxel")()$reference`
* Description: `r readModelDb("Koolen_2010_docetaxel")()$description`
* Article: [Br J Clin Pharmacol 2010;69(5):465-474](https://doi.org/10.1111/j.1365-2125.2010.03621.x)

This is a joint popPK model for intravenous and oral docetaxel, with concomitant
oral ritonavir as a CYP3A4 perpetrator that boosts docetaxel exposure both by
increasing gut bioavailability (`F_gut` rises from 19% to 39% in the presence of
ritonavir) and by competitively inhibiting docetaxel intrinsic clearance via a
well-stirred hepatic-extraction model (Wilkinson 1975). The ritonavir PK is
embedded as a one-compartment first-order-absorption model whose typical
parameters are fixed at the Kappelhoff 2005 ritonavir popPK estimates
(`modellib("Kappelhoff_2005_ritonavir")`); ritonavir was not re-estimated in
Koolen 2010, only its individual Bayesian posthoc estimates were used. The
docetaxel disposition is the Bruno 1996 three-compartment IV model with a
single-transit absorption chain added for the oral route.

# Population

The model was fit to pooled data from two single-centre phase I studies
conducted at the Netherlands Cancer Institute / Slotervaart Hospital. Thirty-six
adults (20 male, 16 female; median age 54 years, range 31-73) with
histologically or cytologically confirmed advanced cancer for whom no standard
therapy was available contributed 72 treatment courses, comprising 1025
docetaxel and 276 ritonavir plasma concentrations sampled at predose, 0.25,
0.5, 0.75, 1, 1.25, 1.5, 2, 3, 4, 6, 8, 10, 24, 36, and 48 h after
administration (Koolen 2010 Methods Patients section and Table 1).

Per-arm patient counts (Koolen 2010 Table 1):

| Regimen                                       | N |
|-----------------------------------------------|---|
| Docetaxel IV 100 mg/m^2                       | 15 |
| Docetaxel IV 100 mg                           | 17 |
| Oral docetaxel 75 mg/m^2 alone                | 3 |
| RTV 100 mg + oral docetaxel 10 mg, 1 h apart  | 6 |
| RTV 100 mg + oral docetaxel 100 mg, 1 h apart | 15 |
| RTV 100 mg + oral docetaxel 10 mg simultaneous | 5 |
| RTV 100 mg + oral docetaxel 100 mg simultaneous | 11 |

The same baseline information is accessible programmatically:

```{r population, eval = FALSE}
readModelDb("Koolen_2010_docetaxel")()$population
```

# Source trace

The per-parameter origin is recorded inline in the model file
(`inst/modeldb/specificDrugs/Koolen_2010_docetaxel.R`). The table below
collects each equation / parameter and its source location for review.

| Equation / parameter        | Value             | Source location |
|-----------------------------|-------------------|------------------|
| `lcli0` (CLi,0)             | 113 L/h           | Table 2, "CL i,0" row, Final model column |
| `lvc_iv` (V_central_iv)     | 9.8 L             | Table 2, "V2 (iv)" row, Final model column |
| `lvc_po` (V_central_po)     | 44.0 L            | Table 2, "V2 (po)" row, Final model column |
| `lq` (Q1)                   | 6.9 L/h           | Table 2, "Q1" row, Final model column |
| `lvp` (V_peripheral1)       | 7.5 L             | Table 2, "V3" row, Final model column |
| `lq2` (Q2)                  | 15.7 L/h          | Table 2, "Q2" row, Final model column |
| `lvp2` (V_peripheral2)      | 376 L             | Table 2, "V4" row, Final model column |
| `lmat` (MAT)                | 1.3 h             | Table 2, "MAT" row, Final model column |
| `qhep` (Q_hep, fixed)       | 80 L/h            | Table 2, "Q (Hepatic blood flow, fixed)" row |
| `lfdepot` (F_doc)           | 0.19              | Table 2, "F1" row, Final model column |
| `e_rtv_fdepot`              | log(0.39/0.19)    | Table 2, "F_RTV" row paired with F1 |
| `lki` (Ki)                  | 0.028 ug/mL       | Table 2, "K i" row, Final model column |
| `lcl_rtv` (CL/F RTV)        | 10.5 L/h FIXED    | Kappelhoff 2005 Table 2 (Koolen 2010 ref [8]) |
| `lvc_rtv` (V/F RTV)         | 96.6 L FIXED      | Kappelhoff 2005 Table 2 (Koolen 2010 ref [8]) |
| `lka_rtv` (ka RTV)          | 0.871 1/h FIXED   | Kappelhoff 2005 Table 2 (Koolen 2010 ref [8]) |
| `ltlag_rtv` (Tlag RTV)      | 0.778 h FIXED     | Kappelhoff 2005 Table 2 (Koolen 2010 ref [8]) |
| `propSd` (proportional RUV) | 0.32              | Table 2, "P (Proportional error)" row, Final model column |
| IIV CLi,0 (eta variance)    | 60% CV            | Table 2, "CL i,0" row, IIV (CV%) column |
| IIV V2 (iv)                 | 45% CV            | Table 2, "V2 (iv)" row, IIV (CV%) column |
| IIV V2 (po)                 | 35% CV            | Table 2, "V2 (po)" row, IIV (CV%) column |
| IIV MAT                     | 87% CV            | Table 2, "MAT" row, IIV (CV%) column |
| IIV F (shared F1, F_RTV)    | 72% CV            | Table 2, "F1" and "F_RTV" rows, IIV (CV%) column |
| IIV Ki                      | 122% CV           | Table 2, "K i" row, IIV (CV%) column |
| Correlation CLi,0 ~ V2 (iv) | 0.446             | Table 2, "CL ~ V2 Correlation" row, Final model column |
| Equation 1 (F_hep)          | Q_hep/(CLi+Q_hep) | Methods page 467, Equation 1; Wilkinson 1975 |
| Equation 2 (CL)             | Q_hep*CLi/(CLi+Q_hep) | Methods page 467, Equation 2 |
| Equation 3 (CLi)            | CLi,0/(1+[RTV]/Ki) | Methods page 467, Equation 3 |
| Equation 4 (F_gut)          | F_doc + (F_RTV-F_doc)*RTV_ind | Methods page 467, Equation 4 (non-RTV-conc-dependent) |
| ODE system (compartments)   | Figure 1 equations | Methods Figure 1, page 467 |
| ktr = 2 / MAT               | ktr = 1.538 1/h   | Figure 1 caption, page 467 |

# Virtual cohort

Original observed data are not publicly available. To validate against Koolen
2010 Table 3 (geometric-mean AUC0-inf for 1000 simulated patients across ten
ritonavir dosing strategies), the chunk below recreates six of the ten regimens
the paper's Table 3 reports: single 50 / 100 / 200 mg ritonavir doses given
either simultaneously or 1 h prior to a single 100 mg oral docetaxel dose. All
covariates are set per dose record so the `V_central_iv` vs `V_central_po`
switch and the `F_doc` vs `F_RTV` switch behave correctly.

```{r cohort}
set.seed(20100101)

n_per_arm <- 200L  # smaller than the paper's 1000 to keep the vignette fast

# One (id, time) row per dose event and per observation time. Each cohort
# uses a disjoint integer ID range so bind_rows() does not silently merge
# subjects across cohorts.
build_arm <- function(n, dose_doc_mg, dose_rtv_mg, rtv_offset_h, label,
                      id_offset = 0L,
                      obs_times = c(0.25, 0.5, 0.75, 1, 1.25, 1.5, 2,
                                    3, 4, 6, 8, 10, 12, 24, 36, 48, 72)) {
  ids <- id_offset + seq_len(n)

  # Docetaxel oral dose at t = 0 into the depot compartment.
  doc_dose <- expand.grid(id = ids, time = 0,
                          stringsAsFactors = FALSE)
  doc_dose$evid       <- 1L
  doc_dose$cmt        <- "depot"
  doc_dose$amt        <- dose_doc_mg
  doc_dose$ROUTE_IV   <- 0L
  doc_dose$CONMED_RTV <- 1L

  # Ritonavir oral dose at t = rtv_offset_h into depot_rtv. The paper's
  # "1 h prior" arms use rtv_offset_h = -1; the "simultaneous" arms use
  # rtv_offset_h = 0.
  rtv_dose <- expand.grid(id = ids, time = rtv_offset_h,
                          stringsAsFactors = FALSE)
  rtv_dose$evid       <- 1L
  rtv_dose$cmt        <- "depot_rtv"
  rtv_dose$amt        <- dose_rtv_mg
  rtv_dose$ROUTE_IV   <- 0L
  rtv_dose$CONMED_RTV <- 1L

  # Observation rows (cmt = NA -> default observation of Cc).
  obs <- expand.grid(id = ids, time = obs_times, stringsAsFactors = FALSE)
  obs$evid       <- 0L
  obs$cmt        <- NA_character_
  obs$amt        <- 0
  obs$ROUTE_IV   <- 0L
  obs$CONMED_RTV <- 1L

  out <- dplyr::bind_rows(doc_dose, rtv_dose, obs)
  out$treatment <- label
  dplyr::arrange(out, id, time)
}

arms <- dplyr::bind_rows(
  build_arm(n_per_arm, 100, 50,   0, "RTV 50 mg simultaneous",   id_offset = 0L * n_per_arm),
  build_arm(n_per_arm, 100, 100,  0, "RTV 100 mg simultaneous",  id_offset = 1L * n_per_arm),
  build_arm(n_per_arm, 100, 200,  0, "RTV 200 mg simultaneous",  id_offset = 2L * n_per_arm),
  build_arm(n_per_arm, 100, 50,  -1, "RTV 50 mg 1 h prior",      id_offset = 3L * n_per_arm),
  build_arm(n_per_arm, 100, 100, -1, "RTV 100 mg 1 h prior",     id_offset = 4L * n_per_arm),
  build_arm(n_per_arm, 100, 200, -1, "RTV 200 mg 1 h prior",     id_offset = 5L * n_per_arm)
)

# Defensive: assert IDs are disjoint across cohorts.
stopifnot(!anyDuplicated(unique(arms[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod <- readModelDb("Koolen_2010_docetaxel")()

sim <- rxode2::rxSolve(mod, events = arms, keep = c("treatment", "ROUTE_IV",
                                                    "CONMED_RTV")) |>
  as.data.frame()
```

# Replicate published figures

## Figure 2 -- docetaxel CL versus time with concomitant ritonavir

Koolen 2010 Figure 2 plots the time course of estimated docetaxel clearance
(L/h) for individuals who received 100 mg oral docetaxel + 100 mg ritonavir
simultaneously, alongside the typical-value ritonavir plasma-concentration
trajectory. The figure shows that ritonavir suppresses docetaxel CL by
approximately 90% almost immediately after dosing, with gradual recovery as
ritonavir declines.

```{r figure-2}
sim_typ <- rxode2::rxSolve(rxode2::zeroRe(mod), events = arms |>
                             dplyr::filter(treatment == "RTV 100 mg simultaneous"),
                           keep = c("treatment")) |>
  as.data.frame()

# Median CL and median crtv across subjects at each time.
cl_curve <- sim_typ |>
  dplyr::filter(time >= 0) |>
  dplyr::group_by(time) |>
  dplyr::summarise(cl_median  = median(cl,   na.rm = TRUE),
                   rtv_median = median(crtv, na.rm = TRUE),
                   .groups    = "drop")

cl_max_unboost <- 80 * 113 / (80 + 113)  # CL when [RTV] = 0

ggplot(cl_curve, aes(time, cl_median)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_hline(yintercept = cl_max_unboost, linetype = "dashed",
             colour = "darkgray") +
  scale_x_continuous(limits = c(0, 24)) +
  scale_y_continuous(limits = c(0, cl_max_unboost * 1.05)) +
  labs(x = "Time (h)", y = "Docetaxel CL (L/h)",
       title = "Figure 2: docetaxel CL after 100 mg oral docetaxel + 100 mg simultaneous ritonavir",
       caption = paste0(
         "Replicates Figure 2 of Koolen 2010. Dashed line is the unboosted ",
         "typical-subject CL = Q_hep * CLi,0 / (CLi,0 + Q_hep) = ",
         round(cl_max_unboost, 1), " L/h."))
```

## Visual predictive check of docetaxel concentrations

```{r vpc-style}
sim |>
  dplyr::filter(time >= 0, time <= 48) |>
  dplyr::group_by(treatment, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc * 1000, 0.05, na.rm = TRUE),  # convert mg/L -> ng/mL
    Q50 = quantile(Cc * 1000, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc * 1000, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  facet_wrap(~ treatment, ncol = 3) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Docetaxel Cc (ng/mL)",
       title = "Simulated 90% prediction band and median",
       caption = "Six of the ten RTV dose-and-timing regimens reported in Koolen 2010 Table 3.")
```

# PKNCA validation

The chunk below derives AUC(0,inf) for each (id, treatment) using PKNCA and
compares the geometric mean across subjects against the values printed in
Koolen 2010 Table 3.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment) |>
  dplyr::mutate(Cc = Cc * 1000)  # convert mg/L -> ng/mL for the NCA output

# Guarantee a time = 0 row per (id, treatment); for extravascular pre-dose
# Cc = 0 is correct.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |>
    dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "ng/mL", timeu = "h")

# Doses: docetaxel dose only (NCA is for docetaxel; RTV is the perpetrator).
dose_df <- arms |>
  dplyr::filter(evid == 1, cmt == "depot") |>
  dplyr::select(id, time, amt, treatment)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "mg")

intervals <- data.frame(
  start       = 0,
  end         = Inf,
  cmax        = TRUE,
  tmax        = TRUE,
  aucinf.obs  = TRUE,
  half.life   = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- PKNCA::pk.nca(nca_data)
```

## Comparison against published Table 3

Koolen 2010 Table 3 reports geometric-mean AUC(0,inf) in mg/L*h units across
1000 simulated patients per regimen, with the 90% confidence interval in
parentheses. The table below converts to ng*h/mL (1 mg/L*h = 1000 ng*h/mL) so
both the simulated and the published columns use the same units used in the
NCA output above.

```{r nca_compare}
geom_mean <- function(x) exp(mean(log(x[x > 0]), na.rm = TRUE))

sim_summary <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD == "aucinf.obs") |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(sim_geomean = geom_mean(PPORRES), .groups = "drop")

# Published values from Koolen 2010 Table 3 (mg/L*h converted to ng*h/mL by
# multiplying by 1000).
published <- tibble::tribble(
  ~treatment,                       ~paper_geomean, ~paper_ci_lo, ~paper_ci_hi,
  "RTV 50 mg simultaneous",           1070,           1000,         1130,
  "RTV 100 mg simultaneous",          1360,           1280,         1450,
  "RTV 200 mg simultaneous",          1710,           1610,         2150,
  "RTV 50 mg 1 h prior",              1870,           1770,         1970,
  "RTV 100 mg 1 h prior",             2560,           2410,         2710,
  "RTV 200 mg 1 h prior",             3320,           3140,         3500
)

cmp <- dplyr::left_join(sim_summary, published, by = "treatment") |>
  dplyr::mutate(
    paper_geomean_pretty = sprintf("%.0f (%.0f, %.0f)",
                                   paper_geomean, paper_ci_lo, paper_ci_hi),
    pct_diff = 100 * (sim_geomean - paper_geomean) / paper_geomean
  ) |>
  dplyr::select(treatment, sim_geomean, paper_geomean_pretty, pct_diff)

knitr::kable(
  cmp,
  col.names = c("Treatment", "Simulated AUC0-inf (ng*h/mL)",
                "Published AUC0-inf (ng*h/mL, geom. mean (90% CI))",
                "% diff (sim vs paper)"),
  digits = 1,
  caption = paste0("Geometric-mean AUC0-inf in 200 simulated patients per arm ",
                   "vs Koolen 2010 Table 3 (1000 simulated patients per arm; ",
                   "1 mg/L*h = 1000 ng*h/mL).")
)
```

# Assumptions and deviations

- **Embedded ritonavir PK is fixed at Kappelhoff 2005 typical values.** Koolen
  2010 used MAP Bayesian posthoc per-subject estimates from the Kappelhoff 2005
  ritonavir popPK model (Koolen 2010 Methods, reference [8]). The packaged
  Koolen 2010 model embeds the Kappelhoff 2005 typical values (CL/F = 10.5 L/h,
  V/F = 96.6 L, ka = 0.871 1/h, Tlag = 0.778 h) as `fixed()` parameters with
  zero IIV so that the joint model is self-contained for simulation. Users who
  need per-subject ritonavir variability should simulate ritonavir
  concentrations from `modellib("Kappelhoff_2005_ritonavir")` independently
  and supply the resulting profile as a time-varying input.
- **Inter-occasion variability (IOV) is not propagated.** Koolen 2010 Table 2
  reports IOV on CLi,0 (22%), MAT (52%), and F_RTV (44%) -- treatment courses
  in the source were three or more weeks apart so per-cycle drift was
  identifiable. The packaged model carries only the inter-individual etas.
  For simulation studies whose endpoint depends on cycle-to-cycle variability,
  the IOV terms can be added on top by hand using `rxode2`'s occasion-block
  facilities.
- **Elevated first-4-hour-after-oral proportional residual error is not
  encoded.** Koolen 2010 reports a 32% proportional RUV for the final model
  plus a separately-estimated 63% proportional RUV for docetaxel observations
  within the first 4 hours after oral administration. Only the structural 32%
  term is propagated by `propSd`. Users who want to overlay a piecewise RUV
  band on a VPC can post-process the simulated `Cc` column by multiplying the
  RUV envelope by 63 / 32 for observation records satisfying
  `ROUTE_IV == 0` and `tad < 4 h`.
- **`ROUTE_IV` reference category is oral, not SC.** The canonical
  `ROUTE_IV` in `inst/references/covariate-columns.md` historically used SC as
  the 0-reference category (Yu 2022, Zierhut 2008, Wang 2021, Fiedler-Kelly
  2019). The Koolen 2010 paper pools IV and oral docetaxel cohorts; the
  packaged model uses `ROUTE_IV = 0` for oral docetaxel and `ROUTE_IV = 1`
  for IV docetaxel, with the V_central switch chosen accordingly. The
  ROUTE_IV register entry now documents this paper-specific reference
  category alongside the SC-reference precedent.
- **Correlation CL ~ V2 is applied to V_central_iv only.** Koolen 2010 Table 2
  reports a single 44.6% correlation between IIV on CL and IIV on V2 in the
  final model. The packaged model treats this as a correlation between the
  etas on CLi,0 and V_central_iv (the V2 that maps to the basic IV-only Bruno
  model carried forward into the final model); V_central_po carries an
  independent eta. The paper does not specify which V2 the correlation belongs
  to; the IV interpretation is the most natural carry-forward from the basic
  IV-only column of Table 2.
- **Vignette uses 200 simulated subjects per arm, not 1000.** Koolen 2010
  Table 3 was generated from 1000 simulated subjects per arm; the validation
  vignette here uses 200 to keep render time under the 5 minute pkgdown gate.
  The geometric-mean point estimates remain a reasonable check; the
  paper's 90% confidence intervals would tighten on larger N.
