---
title: "Factorix (Koopman 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Factorix (Koopman 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Koopman SF, Goedhart TMHJ, Bukkems LH, et al. A new population
  pharmacokinetic model for recombinant factor IX-Fc fusion concentrate
  including young children with haemophilia B. *Br J Clin Pharmacol.*
  2024;90(1):220-231.
  doi:[10.1111/bcp.15881](https://doi.org/10.1111/bcp.15881)
* Description: Two-compartment population PK model for recombinant factor
  IX-Fc fusion protein (rFIX-Fc, eftrenonacog alfa, Alprolix) in haemophilia B
  patients aged 2-71 years.
* Article: <https://doi.org/10.1111/bcp.15881>
* Modality: Fc-fusion factor concentrate, IV bolus.

rFIX-Fc is an extended half-life FIX concentrate consisting of a single
recombinant factor IX molecule fused to the dimeric Fc domain of human IgG1.
The Fc domain binds FcRn, delaying lysosomal degradation and prolonging the
circulation half-life relative to standard rFIX. Koopman 2023 externally
evaluated the only previously published rFIX-Fc population PK model (Diao 2014,
based on patients aged >= 12 years) using real-world data and found systematic
underprediction (median PE -48.8% across all patients, -54.1% in children
<12 years). They then developed a new two-compartment model fit to the
combined OPTI-CLOT TARGET (Netherlands) and UK-EHL Outcomes Registry cohorts.

The packaged model implements the **new Koopman 2023 model** (the right-hand
column of Koopman 2023 Table 2; equations on p. 226), not the published Diao
2014 model. The new model has a single covariate effect: a linear-deviation
slope of age on CL, centered at the cohort median age of 15.8 years.
Allometric scaling of all PK parameters by body weight (reference 73 kg) is
applied with theoretical exponents (0.75 on CL and Q; 1.00 on V1 and V2).

# Population

The development cohort comprised **37 patients with haemophilia B** treated
prophylactically with rFIX-Fc (Koopman 2023 Table 1):

- 35 severe (FIX < 1 IU/dL) and 2 moderately severe haemophilia B.
- Age 2-71 years (median 15.8 years, IQR 11-30).
- Body weight 12-103 kg (median 65.4 kg, IQR 33-77).
- 19/37 (51%) paediatric (< 18 years), 14/37 (38%) < 12 years, 7/37 (19%)
  < 6 years.
- Median dose 36 IU/kg rFIX-Fc IV (range 10-132 IU/kg).
- Median 5 FIX activity samples per PK profile (range 3-7); 287 measurements
  total, 3 below LLOQ excluded.
- Sex: hemophilia B is X-linked, so the cohort is essentially all male.
- Region: Netherlands (OPTI-CLOT TARGET, NTR7523) and United Kingdom (UK-EHL
  Outcomes Registry, NCT02938156).

The same metadata is available programmatically via
`readModelDb("Koopman_2023_factorix")$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in
`inst/modeldb/specificDrugs/Koopman_2023_factorix.R`. The table below collects
them in one place for review.

| Parameter (model name)             | Value                  | Source                                                              |
|------------------------------------|------------------------|---------------------------------------------------------------------|
| `lcl` (typical CL, dL/h)           | log(1.41)              | Koopman 2023 Table 2 (New): CL = 1.41 dL/h                          |
| `lvc` (typical V1, dL)             | log(73.1)              | Koopman 2023 Table 2 (New): V1 = 73.1 dL                            |
| `lq`  (typical Q2, dL/h)           | log(2.77)              | Koopman 2023 Table 2 (New): Q2 = 2.77 dL/h                          |
| `lvp` (typical V2, dL)             | log(80.1)              | Koopman 2023 Table 2 (New): V2 = 80.1 dL                            |
| Allometric exponent on CL, Q       | 0.75 (fixed)           | Koopman 2023 Table 2 (New): bodyweight exponent on CL and Q2 = 0.75 |
| Allometric exponent on V1, V2      | 1.00 (fixed)           | Koopman 2023 Table 2 (New): bodyweight exponent on V1 and V2 = 1.00 |
| `e_age_cl` (linear age slope, 1/y) | 0.0047                 | Koopman 2023 Table 2 (New): age exponent on CL = 0.0047             |
| Reference body weight              | 73 kg                  | Koopman 2023 Table 2 footnote                                       |
| Reference age                      | 15.8 years             | Koopman 2023 equations on p. 226: `(AGE - 15.8)`                    |
| IIV block `etalcl + etalvc`        | c(0.05570, 0.03281, 0.09986) | Koopman 2023 Table 2: IIV CL = 23.6%, IIV V1 = 31.6%, corr CL:V1 = 44.0% |
| `etalvp`                           | 0.16974                | Koopman 2023 Table 2: IIV V2 = 41.2%                                |
| `propSd`                           | 0.163                  | Koopman 2023 Table 2: proportional error = 16.3%                    |
| `addSd`                            | 1.04 IU/dL             | Koopman 2023 Table 2: additive error = 1.04 IU/dL                   |
| Equation: `d/dt(central)`          | n/a                    | Koopman 2023 p. 226 (new model equations): two-compartment IV       |
| Equation: `d/dt(peripheral1)`      | n/a                    | Koopman 2023 p. 226 (new model equations)                           |

The footnote of Koopman 2023 Table 2 states "IIV and IOV coefficient of
variation calculated as: sqrt(variance)\*100%", i.e., the reported CV% values
are `sqrt(omega^2) * 100`, so `omega^2 = (CV/100)^2`. The IIV variances above
were computed by squaring `0.236`, `0.316`, and `0.412`, and the CL:V1
covariance as `0.44 * 0.236 * 0.316`.

The new model also reports inter-occasion variability on CL (IOV CL = 19.8%)
which is not implemented in this static model file; see *Assumptions and
deviations*.

# Virtual cohort

Original observed FIX activity data are not publicly available. The
simulations below use a virtual cohort whose demographics approximate the
Koopman 2023 development population, stratified into the three age groups
used in the paper's half-life table (Supplementary Table 3): children
< 12 years, adolescents 12 - < 18 years, and adults >= 18 years. Body weight
is sampled from a log-normal distribution centered on a typical weight-for-age
within each band, capped at the observed range from Koopman 2023 Table 1.

```{r cohort}
set.seed(2023)

make_cohort <- function(n, age_min, age_max, wt_log_mean, wt_log_sd,
                        wt_lo, wt_hi, label, id_offset = 0L) {
  tibble(
    ID        = id_offset + seq_len(n),
    AGE       = runif(n, age_min, age_max),
    WT        = pmin(pmax(rlnorm(n, log(wt_log_mean), wt_log_sd), wt_lo), wt_hi),
    age_group = label
  )
}

n_per_group <- 100L
cohort <- bind_rows(
  make_cohort(n_per_group, 2,  12, 22, 0.30, 12, 45,
              "Children <12 y",         id_offset = 0L),
  make_cohort(n_per_group, 12, 18, 50, 0.20, 35, 80,
              "Adolescents 12-<18 y",   id_offset = 100L),
  # Adults are sampled over 18-50 to approximate the median age of the
  # development cohort (overall median 15.8 y, paediatric n = 19, adult
  # n = 18, so the adult median is likely in the late twenties / early
  # thirties — a uniform 18-71 sample shifts the median half-life upward
  # because of the linear age-on-CL effect).
  make_cohort(n_per_group, 18, 50, 75, 0.20, 50, 110,
              "Adults >=18 y",          id_offset = 200L)
)

stopifnot(!anyDuplicated(cohort$ID))
summary(cohort)
```

A single 50 IU/kg IV bolus dose of rFIX-Fc is administered at time 0 and FIX
activity is observed over a 14-day window (a typical clinical follow-up
between Q1W prophylaxis doses).

```{r events}
obs_grid <- c(0, 0.25, 0.5, 1, 2, 4, 8, 12, 24,
              seq(48, 336, by = 12))

build_events <- function(pop) {
  amt_per_subject <- pop$WT * 50
  d_dose <- pop |>
    mutate(AMT = amt_per_subject) |>
    tidyr::crossing(TIME = 0) |>
    mutate(EVID = 1, CMT = "central", DV = NA_real_)
  d_obs <- pop |>
    tidyr::crossing(TIME = obs_grid) |>
    mutate(AMT = NA_real_, EVID = 0, CMT = "central", DV = NA_real_)
  bind_rows(d_dose, d_obs) |>
    arrange(ID, TIME, desc(EVID)) |>
    as.data.frame()
}

events <- build_events(cohort)
```

# Simulation

Run a stochastic VPC-style simulation (between-subject variability on CL, V1,
V2 included) and a typical-value simulation with the etas zeroed for direct
parameter back-checks.

```{r simulate}
mod <- readModelDb("Koopman_2023_factorix")

sim <- rxode2::rxSolve(mod, events = events,
                       keep = c("age_group"),
                       returnType = "data.frame")
sim <- sim[sim$time >= 0, ]

mod_typ <- rxode2::zeroRe(mod)
sim_typ <- rxode2::rxSolve(mod_typ, events = events,
                           keep = c("age_group"),
                           returnType = "data.frame")
sim_typ <- sim_typ[sim_typ$time >= 0, ]
```

# FIX activity-time profiles by age group

Koopman 2023 Figure 3 shows the visual predictive check of the new model
across the development cohort. The figure below reproduces the typical-value
median + 5 - 95% prediction interval of FIX activity vs. time after dose,
stratified by the three age bands used in the paper's half-life summary
(Supplementary Table 3). At later times the paediatric profile sits below the
adult profile because (a) typical V1 scales linearly with WT, so per-dose Cmax
is similar across age groups (`Cmax = dose / V1 = (50 * WT) / (73.1 * WT/73)`,
WT-independent), but (b) typical CL/V1 (the apparent first-order elimination
rate constant of the central compartment) is higher at low WT under
allometric scaling with exponent 0.75 / 1.0 (`CL/V1` scales as `WT^(-0.25)`),
so trough activity is lower and terminal half-life is shorter in younger /
smaller patients.

```{r profile-plot, fig.width = 8, fig.height = 5}
sim_summary <- sim |>
  filter(time > 0) |>
  group_by(time, age_group) |>
  summarise(
    median = stats::median(Cc, na.rm = TRUE),
    lo     = stats::quantile(Cc, 0.05, na.rm = TRUE),
    hi     = stats::quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(sim_summary, aes(time, median, colour = age_group, fill = age_group)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.18, colour = NA) +
  geom_line(linewidth = 1) +
  scale_y_log10() +
  labs(
    x        = "Time after dose (h)",
    y        = "FIX activity above baseline (IU/dL, log scale)",
    colour   = "Age group",
    fill     = "Age group",
    title    = "Simulated rFIX-Fc PK after a single 50 IU/kg IV dose",
    subtitle = paste0("Median + 5-95% prediction interval (N = ", n_per_group,
                      " per group). Approximate analogue of Koopman 2023 Figure 3."),
    caption  = "Model: Koopman 2023 Br J Clin Pharmacol 90(1):220-231"
  ) +
  theme_bw()
```

# Typical CL and Vss back-check

Koopman 2023 reports that for a typical 16-year-old, 73 kg patient (the
"reference" subject of the new model), CL = 1.41 dL/h and steady-state volume
of distribution Vss = V1 + V2 = 73.1 + 80.1 = 153 dL. Reproducing those
numbers from the typical-value simulation is the simplest possible
self-consistency check.

```{r typical-cl-vss}
mod_typ <- rxode2::zeroRe(mod)
ev_ref <- rxode2::et(amt = 50 * 73, time = 0, cmt = "central") |>
  rxode2::et(seq(0, 0))
sim_ref <- rxode2::rxSolve(
  mod_typ, events = ev_ref,
  params = data.frame(WT = 73, AGE = 15.8),
  returnType = "data.frame"
)

# rxSolve carries the derived individual parameters as columns when the model
# defines them (cl, vc, q, vp). We pull them out of the first observation row.
ref_pars <- sim_ref[1, c("cl", "vc", "q", "vp"), drop = FALSE]
ref_pars$Vss <- ref_pars$vc + ref_pars$vp

knitr::kable(
  ref_pars,
  caption = "Typical-value PK parameters for the reference 73 kg, 15.8-year-old patient",
  digits  = c(3, 2, 3, 2, 2)
)
```

The model returns CL = `r round(ref_pars$cl, 2)` dL/h, V1 = `r round(ref_pars$vc, 1)` dL,
Q2 = `r round(ref_pars$q, 2)` dL/h, V2 = `r round(ref_pars$vp, 1)` dL, and
Vss = V1 + V2 = `r round(ref_pars$Vss, 1)` dL — matching the values reported
in Koopman 2023 Table 2 (CL = 1.41 dL/h, V1 = 73.1 dL, Q2 = 2.77 dL/h,
V2 = 80.1 dL, Vss = 153 dL).

# PKNCA validation

Use PKNCA to compute Cmax, AUC0-inf and terminal half-life by age group, and
compare the simulated half-lives against Koopman 2023 Supplementary Table 3.

```{r pknca}
sim_nca <- sim |>
  filter(!is.na(Cc), Cc > 0) |>
  select(id, time, Cc, age_group)

dose_df <- events |>
  filter(EVID == 1) |>
  transmute(id = ID, time = TIME, amt = AMT, age_group)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | age_group + id,
                             concu = "IU/dL",
                             timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | age_group + id,
                             doseu = "IU")

intervals <- data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE,
  clast.obs  = TRUE,
  lambda.z   = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
nca_tbl <- as.data.frame(nca_res$result)

half_life_summary <- nca_tbl |>
  filter(PPTESTCD == "half.life") |>
  group_by(age_group) |>
  summarise(
    n        = sum(!is.na(PPORRES)),
    median_h = stats::median(PPORRES, na.rm = TRUE),
    q05      = stats::quantile(PPORRES, 0.05, na.rm = TRUE),
    q95      = stats::quantile(PPORRES, 0.95, na.rm = TRUE),
    .groups  = "drop"
  )

knitr::kable(
  half_life_summary,
  caption = "Simulated rFIX-Fc terminal half-life (h) by age group, single 50 IU/kg IV dose.",
  digits  = c(0, 0, 1, 1, 1)
)
```

## Comparison against Koopman 2023 Supplementary Table 3

Koopman 2023 Supplementary Table 3 reports post-hoc terminal half-lives from
the new model in three age bands. The simulated medians below should fall
within the reported observed ranges; differences > 20% of the reported median
would indicate a coding problem.

```{r half-life-comparison}
published <- tibble::tribble(
  ~age_group,                 ~published_median_h, ~published_range,
  "Children <12 y",           70,                  "51-103",
  "Adolescents 12-<18 y",     76,                  "66-82",
  "Adults >=18 y",            88,                  "67-166"
)

comparison <- published |>
  left_join(
    half_life_summary |>
      select(age_group, simulated_median_h = median_h),
    by = "age_group"
  ) |>
  mutate(
    pct_diff = round(100 * (simulated_median_h - published_median_h) /
                       published_median_h, 1)
  )

knitr::kable(
  comparison,
  caption = "Simulated vs. Koopman 2023 Supplementary Table 3 terminal half-lives.",
  digits  = c(0, 0, 0, 1, 1)
)
```

A typical Cmax check: the typical-value simulation gives the same Cmax in
every age group (`50 * WT / (73.1 * WT / 73)` = 49.93 IU/dL), as expected from
the linear allometric scaling with exponent 1.0 on V1.

# Errata

No published erratum was located for Koopman 2023 (Br J Clin Pharmacol
2024;90(1):220-231). The packaged parameter values are taken from Koopman 2023
Table 2 (the "New" column) and the equations on p. 226, which are internally
consistent: `CL = 1.41 dL/h` is the typical CL at the reference covariates
`WT = 73 kg` and `AGE = 15.8 years`, and Vss = V1 + V2 = 73.1 + 80.1 = 153 dL
matches the discussion text ("distribution volume at steady-state was also
lower with respective values of 153 and 198 dL").

One narrative passage on page 225 appears inconsistent with the equation on
page 226 and Table 2: *"On basis of this relationship, typical clearance of a
73-kg patient would decrease from 1.89 dL/h at age 20 years to 1.36 dL/h at 70
years."* Substituting `(WT = 73, AGE = 20, AGE = 70)` into the published
equation `CL = 1.41 * (73/73)^0.75 * (1 - 0.0047 * (AGE - 15.8))` gives 1.38
dL/h at age 20 and 1.05 dL/h at age 70 - in the same direction (CL decreases
with age) but ~25-35% lower in absolute magnitude than the values quoted in
the discussion. The packaged model uses the Table 2 values and the published
equation as authoritative, since they are internally consistent and reproduce
the half-life table (Supplementary Table 3) within the reported observed
ranges.

# Assumptions and deviations

- **Inter-occasion variability omitted.** Koopman 2023 estimated IOV on CL
  (19.8% CV, RSE 22%, shrinkage 36%) reflecting within-subject variability
  across PK profiling visits. The static library model has no occasion
  variable, so IOV is not implemented as a separate eta. For Bayesian
  forecasting use cases that explicitly model occasions, the IOV variance
  (`omega^2_IOV_CL = 0.198^2 = 0.03920`) should be added on top of the
  packaged IIV.
- **Inter-individual variability convention.** Koopman 2023 footnotes Table 2
  with "IIV and IOV coefficient of variation calculated as: sqrt(variance) \*
  100%", i.e., the reported CV% values are `sqrt(omega^2) * 100`, not the
  log-normal `sqrt(exp(omega^2) - 1)`. The packaged variances were computed
  with the simpler `omega^2 = (CV/100)^2` formula to match the paper's
  convention.
- **No IIV on Q2.** Koopman 2023 Table 2 reports IIV only on CL, V1 and V2;
  no IIV was estimated for Q2. The packaged model carries no `etalq`.
- **FIX activity is baseline-corrected.** The one-stage FIX activity assay
  cannot distinguish endogenous baseline FIX from exogenous rFIX-Fc, so
  Koopman 2023 corrects observations for endogenous baseline and residual
  decay of any prior factor product (their Equations 1 - 3) before fitting.
  The packaged model output `Cc` represents the rFIX-Fc-attributable FIX
  activity above baseline, in IU/dL.
- **Allometric exponents fixed at theoretical values.** Koopman 2023 fixed
  the body weight exponents at 0.75 (CL, Q) and 1.00 (V1, V2) per Anderson &
  Holford 2008, rather than estimating them. The packaged `model()` block
  hard-codes these constants accordingly.
- **Reference age centering.** The age effect is centered at 15.8 years (the
  median age of the development cohort). Setting `AGE = 15.8` recovers the
  typical-value parameters in Table 2; setting `AGE` outside the studied
  range (2 - 71 years) extrapolates the linear age slope without empirical
  support.
- **Virtual cohort.** Demographics were simulated to span the three age bands
  used in Koopman 2023 Supplementary Table 3 (children < 12 y, adolescents
  12 - < 18 y, adults >= 18 y) with weight-for-age sampled from a log-normal
  distribution capped at the observed weight range from Koopman 2023 Table 1
  (12-103 kg). Joint covariate structure (e.g., narrow weight range within
  each pediatric age year) is not simulated.
- **Number of compartments.** Koopman 2023 fit a 2-compartment model and
  noted that the rapid distribution phase (occurring within 2-3 h of the end
  of administration, captured by the third compartment in Diao 2014) was not
  resolvable from their sparser real-world sampling schedule. Cmax and trough
  predictions for sampling beyond ~4 h post-dose are not affected; users who
  need to model the very-early distribution phase should consider Diao 2014
  instead.
- **Hemophilia B is X-linked recessive**, so the development cohort and the
  virtual cohort here are essentially all male (`sex_female_pct = 0` in the
  population metadata). The model has no sex covariate.
