---
title: "Daridorexant (Krause 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Daridorexant (Krause 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Krause_2023_daridorexant"))$reference`
* Description: `r rxode2::rxode(readModelDb("Krause_2023_daridorexant"))$description`
* Article (open access): <https://doi.org/10.1002/psp4.12877>

Daridorexant is a dual orexin receptor antagonist for insomnia, dosed at
bedtime. Krause 2023 pooled 13 phase 1, two phase 2 and two phase 3 studies
into one Monolix (SAEM) population PK model: two compartments, first-order
absorption after a lag time, Michaelis-Menten elimination and a relative
bioavailability that falls with dose. The paper reports one final model, so the
package carries one model file.

# Population

The phase 1 studies contributed 9420 concentrations from 412 subjects with
dense sampling (healthy subjects and subjects with hepatic or renal
impairment); the two phase 2 and two phase 3 studies contributed 3207
next-morning concentrations from 1486 patients with insomnia disorder, 12 627
concentrations from 1898 subjects in total (Results, Data characteristics).
Subjects were 18-88 years old (median 54), weighed 42-119.5 kg (median 73.6),
had a lean body weight of 29.1-83.9 kg (median 47.2) and a fat mass of 7.0-58.6
kg (median 23.4); 1155 were female and 1646 White (Table 1). Oral doses ranged
from 5 to 200 mg. Phase 1 subjects were mostly dosed in the morning, phase 2
and 3 patients at bedtime.

```{r population}
str(rxode2::rxode(readModelDb("Krause_2023_daridorexant"))$population)
```

# Source trace

All values are from Table 2, column "All data, final model". Parameters
without an RSE in that column were held fixed at the phase 1 estimates (Methods
step 4 and the Table 2 note) and carry `fixed()` in the model file.

| Item | Value | Source location |
|---|---|---|
| Structure: 2 compartments, first-order absorption with lag, Michaelis-Menten elimination | depot, central, peripheral1 | Abstract; Results, Population PK model |
| Covariate forms: `theta * (cov/ref)^beta` (continuous), `theta * exp(beta)` (categorical) | -- | Methods, Model development equations |
| References: age 30 y, fat mass 20 kg, lean body weight 55 kg, ALP 60 U/L, dose 50 mg, light meal / uncontrolled, evening dosing | -- | Table 2 footnote |
| `lfdepot` F | 0.41 (fixed) | Table 2 'F' |
| `e_dose_fdepot` | -0.47 (fixed) | Table 2 'Dose' |
| `ltlag` | 0.41 h (fixed) | Table 2 't lag (h)' |
| `e_fasted_tlag`, `e_fed_highfat_tlag` | -0.51, 0.62 (fixed) | Table 2 'Food: fasted / fed on t lag' |
| `lka` | 1.05 1/h (fixed) | Table 2 'k a (1/h)' |
| `e_fasted_ka`, `e_fed_highfat_ka`, `e_morning_ka` | 0.09, -1.19, 1.05 (fixed) | Table 2 food and 'Morning administration on k a' rows |
| `lvc`, `e_ffm_vc` | 14.6 L (fixed), 0.49 | Table 2 'V c', 'Lean body weight on V c' |
| `lvp`, `e_fm_vp` | 13.7 L (fixed), 0.47 | Table 2 'V p', 'Fat mass on V p' |
| `lq`, `e_fm_q` | 3.58 L/h (fixed), 1.47 | Table 2 'Q', 'Fat mass on Q' |
| `lvmax` | 6.94 mg/h (fixed) | Table 2 'V m (mg/h)' |
| `lkm` | 2.36 ug/mL (fixed) | Table 2 'K m (ug/ml)' |
| `e_age_km`, `e_alp_km`, `e_ffm_km` | 0.30, 0.34, -0.20 | Table 2 'Age / ALP / Lean body weight on K m' |
| IIV SD on F, tlag, ka, Vc, Vp, Q, Vm, Km | 0.42, 0.26, 0.59, 0.17, 0.23, 0.50, 0.09, 0.36 (fixed; variance = SD^2) | Table 2 'SD(.)' rows |
| IOV SD on F, tlag, ka | 0.20, 0.42, 0.66 (fixed; variance = SD^2) | Table 2 'IOV(.)' rows |
| `addSd`, `propSd` | 17.59 ng/mL, 0.16 | Table 2 'Additive error', 'Multiplicative error' |

# Typical-value replication of Figure 3

Figure 3 of Krause 2023 shows model-predicted day-1 exposure after a single 50
mg dose for a reference subject (age 60 years, fat mass 20 kg, lean body weight
55 kg, ALP 60 U/L, light meal / uncontrolled, bedtime dosing) and for the 5th
and 95th percentiles of each continuous covariate and each food / dosing-time
category, varied one at a time. The same scenarios are simulated here with all
random effects set to zero.

```{r fig3-scenarios}
mod <- readModelDb("Krause_2023_daridorexant")
mod_typ <- rxode2::zeroRe(mod)

ref_cov <- data.frame(
  DOSE = 50, AGE = 60, FM = 20, FFM = 55, ALP = 60,
  FED = 1, FED_HIGHFAT = 0, DOSETIME_EVENING = 1, OCC = 1
)
scen_def <- list(
  "Reference" = list(),
  "Age 23 years" = list(AGE = 23),
  "Age 74 years" = list(AGE = 74),
  "Fat mass 13.1 kg" = list(FM = 13.1),
  "Fat mass 39.1 kg" = list(FM = 39.1),
  "Lean body weight 36.2 kg" = list(FFM = 36.2),
  "Lean body weight 69.1 kg" = list(FFM = 69.1),
  "ALP 38 U/L" = list(ALP = 38),
  "ALP 99 U/L" = list(ALP = 99),
  "Fasted" = list(FED = 0),
  "High-fat meal" = list(FED_HIGHFAT = 1),
  "Morning dosing" = list(DOSETIME_EVENING = 0)
)

scen_events <- function(i, nm, dose = 50, n_days = 1) {
  cov <- ref_cov
  for (v in names(scen_def[[nm]])) cov[[v]] <- scen_def[[nm]][[v]]
  cov$DOSE <- dose
  doses <- data.frame(time = 24 * (seq_len(n_days) - 1), amt = dose, evid = 1L, cmt = "depot")
  obs <- data.frame(
    time = sort(unique(c(seq(0, 24 * n_days, by = 0.05), 8))),
    amt = 0, evid = 0L, cmt = "central"
  )
  ev <- dplyr::bind_rows(doses, obs) |> dplyr::arrange(time, dplyr::desc(evid))
  ev$id <- i
  ev$scenario <- nm
  cbind(ev, cov[rep(1, nrow(ev)), , drop = FALSE])
}

ev_fig3 <- dplyr::bind_rows(lapply(seq_along(scen_def), function(i) scen_events(i, names(scen_def)[i])))
sim_fig3 <- rxode2::rxSolve(mod_typ, ev_fig3, keep = "scenario", returnType = "data.frame")
sim_fig3 <- sim_fig3[!is.na(sim_fig3$Cc), ]
```

Exposure metrics are derived with PKNCA (AUC over 0-24 h, Cmax, tmax); the
concentration at 8 h is read off the simulation grid.

```{r fig3-pknca}
conc_obj <- PKNCA::PKNCAconc(
  dplyr::filter(sim_fig3, !is.na(Cc)),
  Cc ~ time | scenario + id
)
dose_df <- ev_fig3 |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, scenario, time, amt)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | scenario + id)
intervals <- data.frame(start = 0, end = 24, cmax = TRUE, tmax = TRUE, auclast = TRUE)
nca_fig3 <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_fig3$result) |>
  dplyr::select(scenario, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
c8 <- sim_fig3 |>
  dplyr::filter(time == 8) |>
  dplyr::select(scenario, c8h = Cc)
fig3_model <- dplyr::left_join(nca_wide, c8, by = "scenario")
fig3_model$scenario <- factor(fig3_model$scenario, levels = names(scen_def))
fig3_model <- dplyr::arrange(fig3_model, scenario)

fig3_model |>
  dplyr::mutate(dplyr::across(c(auclast, cmax, c8h), ~ signif(.x, 4)), tmax = round(tmax, 2)) |>
  dplyr::rename(
    "Scenario" = scenario, "AUC0-24 (ng*h/mL)" = auclast, "Cmax (ng/mL)" = cmax,
    "tmax (h)" = tmax, "C8h (ng/mL)" = c8h
  ) |>
  knitr::kable(caption = "Typical-value day-1 exposure after a single 50 mg dose (compare Figure 3a of Krause 2023).")
```

The Results text quotes the reference subject and the two largest effects:
Cmax 817 ng/mL and tmax 2.1 h for the reference subject, Cmax 523 ng/mL and
tmax 4.5 h with a high-fat meal, and Cmax 1054 ng/mL and tmax 1.25 h with
morning dosing. The remaining Figure 3a end points were digitised by the
maintainers from the published figure: the Cmax and tmax axes were calibrated
on the values quoted in the text, and the AUC0-24 and C8h axes on their tick
marks. Reading precision is about 1% of each axis span.

```{r fig3-compare}
paper_fig3 <- data.frame(
  scenario = names(scen_def),
  auclast = c(7930, 6665, 8214, 8266, 7340, 8677, 7537, 7240, 8703, 7967, 7400, 8084),
  cmax = c(817, 790, 823, 899, 647, 947, 752, 803, 832, 840, 523, 1054),
  c8h = c(382, 317, 395, 409, 365, 405, 368, 347, 417, 372, 443, 360),
  tmax = c(2.1, NA, NA, 2.30, 1.82, NA, NA, NA, NA, 1.84, 4.5, 1.25)
)
sim_long <- as.data.frame(nca_fig3$result) |>
  dplyr::select(scenario, PPTESTCD, PPORRES)
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_long,
  reference = dplyr::select(paper_fig3, -c8h),
  by = "scenario",
  units = c(cmax = "ng/mL", tmax = "h", auclast = "ng*h/mL"),
  tolerance_pct = 5
)
knitr::kable(cmp, caption = "Simulated versus published typical-value day-1 exposure after 50 mg (Krause 2023 Results and Figure 3a).")

c8_cmp <- dplyr::left_join(
  dplyr::select(paper_fig3, scenario, Reference = c8h),
  dplyr::mutate(dplyr::select(fig3_model, scenario, Simulated = c8h), scenario = as.character(scenario)),
  by = "scenario"
) |>
  dplyr::mutate(Simulated = signif(Simulated, 4), pct_diff = round(100 * (Simulated / Reference - 1), 1))
c8_cmp |>
  dplyr::rename("Scenario" = scenario, "Figure 3a" = Reference, "% diff" = pct_diff) |>
  knitr::kable(caption = "Concentration 8 h after a 50 mg bedtime dose (ng/mL), simulated versus Figure 3a.")

fig3_chk <- dplyr::left_join(
  tidyr::pivot_longer(paper_fig3, -scenario, names_to = "metric", values_to = "paper"),
  dplyr::mutate(fig3_model, scenario = as.character(scenario)) |>
    tidyr::pivot_longer(-scenario, names_to = "metric", values_to = "model"),
  by = c("scenario", "metric")
) |>
  dplyr::filter(!is.na(paper)) |>
  dplyr::mutate(pct_diff = 100 * (model / paper - 1))
stopifnot(
  # Typical-value solves with no random effects, so the only discrepancies are
  # the digitising precision and the paper's own simulation grid. A wrong
  # parameter, unit or covariate reference shifts these by tens of percent
  # (an age reference of 60 instead of 30 years lowers AUC0-24 by 12%).
  nrow(fig3_chk) == 42,
  abs(stats::median(fig3_chk$pct_diff)) < 1,
  max(abs(fig3_chk$pct_diff)) < 3
)
```

```{r fig3-forest, fig.height = 6, fig.cap = "Replicates Figure 3b of Krause 2023: percent change of each exposure metric from the reference subject, one covariate at a time."}
ref_row <- fig3_model[fig3_model$scenario == "Reference", ]
fig3_model |>
  dplyr::filter(scenario != "Reference") |>
  tidyr::pivot_longer(c(auclast, cmax, c8h, tmax), names_to = "metric", values_to = "value") |>
  dplyr::mutate(
    ref = c(auclast = ref_row$auclast, cmax = ref_row$cmax, c8h = ref_row$c8h, tmax = ref_row$tmax)[metric],
    pct = 100 * (value / ref - 1),
    metric = factor(metric,
      levels = c("auclast", "cmax", "c8h", "tmax"),
      labels = c("AUC0-24", "Cmax", "C8h", "tmax")
    )
  ) |>
  ggplot2::ggplot(ggplot2::aes(pct, scenario)) +
  ggplot2::geom_vline(xintercept = 0, linetype = 2) +
  ggplot2::geom_point() +
  ggplot2::facet_wrap(~metric, nrow = 1) +
  ggplot2::labs(x = "Percent change from reference subject", y = NULL) +
  ggplot2::theme_bw()
```

The food and dosing-time effects match the Results (high-fat meal: Cmax -36%,
tmax 2.1 to 4.5 h; morning dosing: Cmax +29%, tmax 2.1 to 1.25 h). The paper
states that all other covariates change exposure by less than 20%; the largest
of them, Cmax at a fat mass of 39.1 kg, is -21% both here and in the published
Figure 3a, so the statement is a rounded summary.

The tmax end points for lean body weight are not compared. The model gives
2.00 h at 36.2 kg and 2.15 h at 69.1 kg, a larger central volume delaying the
peak; Figure 3a places the same two end points at about 2.00 h and 2.18 h but
labels them the other way round, which would need the central volume to shrink
with lean body weight, contrary to Table 2 and Table 3 and to the Cmax panel of
the same figure.

# Concentration-time profiles (Figure 4)

```{r fig4, fig.cap = "Replicates Figure 4 of Krause 2023: typical-value profiles with 50 mg once daily at bedtime for 3 days, one covariate varied at a time."}
ev_fig4 <- dplyr::bind_rows(lapply(seq_along(scen_def), function(i) scen_events(i, names(scen_def)[i], n_days = 3)))
sim_fig4 <- rxode2::rxSolve(mod_typ, ev_fig4, keep = "scenario", returnType = "data.frame")
sim_fig4$scenario <- factor(sim_fig4$scenario, levels = names(scen_def))
ggplot2::ggplot(sim_fig4, ggplot2::aes(time, Cc, colour = scenario)) +
  ggplot2::geom_line() +
  ggplot2::labs(x = "Time (h)", y = "Daridorexant (ng/mL)", colour = NULL) +
  ggplot2::theme_bw()
```

The reference-subject curve of Figure 4 (green, "Age: 60") was digitised by
the maintainers at times away from the steep absorption phase. The model
follows it, including the modest rise of the trough over three days (from
about 120 to 150 ng/mL) that the Michaelis-Menten elimination produces:

```{r fig4-compare}
fig4_paper <- data.frame(
  time = c(8, 12, 16, 20, 23.5, 32, 40, 47.5, 56, 64, 71.5),
  paper = c(378, 274, 207, 157, 122, 462, 257, 152, 484, 270, 160)
)
ref4 <- sim_fig4[sim_fig4$scenario == "Reference", ]
fig4_paper$model <- stats::approx(ref4$time, ref4$Cc, xout = fig4_paper$time)$y
fig4_paper$pct_diff <- round(100 * (fig4_paper$model / fig4_paper$paper - 1), 1)
fig4_paper |>
  dplyr::mutate(model = signif(model, 4)) |>
  dplyr::rename(
    "Time (h)" = time, "Figure 4 (ng/mL)" = paper, "Model (ng/mL)" = model,
    "% diff" = pct_diff
  ) |>
  knitr::kable(caption = "Reference-subject concentrations, 50 mg once daily at bedtime, model versus Figure 4.")
stopifnot(
  # Typical-value solve; the tolerance covers digitising precision only.
  abs(stats::median(fig4_paper$pct_diff)) < 2,
  max(abs(fig4_paper$pct_diff)) < 5
)
```

# Dose-dependent bioavailability and nonlinear elimination

Relative bioavailability follows `F = 0.41 * (DOSE / 50)^-0.47`, so a 200 mg
dose has 52% of the bioavailability of a 50 mg dose (Results; Table 3). The
apparent clearance of the Michaelis-Menten term is `Vm / (Km + C)`. For the
Table 2 reference values (Km = 2.36 ug/mL) it is 2.43 L/h at 500 ng/mL, the
"2.4 L/h at concentrations of 500 ng/ml" of the Discussion.

```{r dose-f}
f_ratio <- (200 / 50)^-0.47
cl_500 <- 6.94 / (2.36 + 0.5)
c(F200_over_F50 = f_ratio, CL_at_500_ng_per_mL = cl_500)
stopifnot(abs(f_ratio - 0.52) < 0.005, abs(cl_500 - 2.4) < 0.05)
```

```{r dose-range, fig.cap = "Typical-value single-dose profiles across the studied dose range (reference subject, bedtime, light meal)."}
doses <- c(5, 25, 50, 100, 200)
ev_dose <- dplyr::bind_rows(lapply(seq_along(doses), function(i) {
  scen_events(i, "Reference", dose = doses[i]) |> dplyr::mutate(scenario = paste(doses[i], "mg"))
}))
sim_dose <- rxode2::rxSolve(mod_typ, ev_dose, keep = "scenario", returnType = "data.frame")
sim_dose$scenario <- factor(sim_dose$scenario, levels = paste(doses, "mg"))
ggplot2::ggplot(sim_dose, ggplot2::aes(time, Cc, colour = scenario)) +
  ggplot2::geom_line() +
  ggplot2::labs(x = "Time after dose (h)", y = "Daridorexant (ng/mL)", colour = "Dose") +
  ggplot2::theme_bw()
```

# Stochastic simulation in a virtual insomnia cohort

A virtual cohort of 200 patients receives 50 mg at bedtime with a light meal /
uncontrolled food intake. Body composition is derived the way the paper did:
lean body weight by the Janmahasatian formula from weight, BMI and sex, and fat
mass as the difference between body weight and lean body weight. Weight,
height, age and ALP distributions are chosen to land near the Table 1 medians;
they are not taken from the paper.

```{r cohort}
rxode2::rxSetSeed(20230101)
n_sub <- 200
sexf <- stats::rbinom(n_sub, 1, 1155 / 1898)
ht_m <- ifelse(sexf == 1, stats::rnorm(n_sub, 1.63, 0.065), stats::rnorm(n_sub, 1.76, 0.07))
wt <- ifelse(sexf == 1, stats::rlnorm(n_sub, log(68), 0.17), stats::rlnorm(n_sub, log(80), 0.15))
wt <- pmin(pmax(wt, 42), 119.5)
bmi <- wt / ht_m^2
ffm <- ifelse(sexf == 1, 9270 * wt / (8780 + 244 * bmi), 9270 * wt / (6680 + 216 * bmi))
cohort <- data.frame(
  id = seq_len(n_sub),
  AGE = round(pmin(pmax(stats::rnorm(n_sub, 52, 16), 18), 88)),
  FFM = ffm,
  FM = wt - ffm,
  ALP = pmin(pmax(stats::rlnorm(n_sub, log(61), 0.3), 16), 244),
  DOSE = 50, FED = 1, FED_HIGHFAT = 0, DOSETIME_EVENING = 1, OCC = 1
)
summary(cohort[, c("AGE", "FFM", "FM", "ALP")])
```

```{r stoch-sim}
obs_grid <- data.frame(time = c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 2.5, 3, 4, 5, 6, 8, 9, 10, 12, 16, 20, 24))
ev_st <- dplyr::bind_rows(
  dplyr::mutate(cohort, time = 0, amt = 50, evid = 1L, cmt = "depot"),
  dplyr::cross_join(cohort, obs_grid) |> dplyr::mutate(amt = 0, evid = 0L, cmt = "central")
) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  # Event columns first: rxode2 reads a DOSE column that precedes amt as the
  # dose amount rather than as a covariate.
  dplyr::select(id, time, amt, evid, cmt, dplyr::everything())
sim_st <- rxode2::rxSolve(mod, ev_st, returnType = "data.frame")
sim_st$treatment <- "50 mg bedtime"
```

```{r vpc, fig.cap = "Stochastic simulation, 50 mg at bedtime: median and 5th-95th percentiles of plasma daridorexant (residual error included)."}
sim_st |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    p05 = stats::quantile(sim, 0.05), p50 = stats::median(sim), p95 = stats::quantile(sim, 0.95),
    .groups = "drop"
  ) |>
  ggplot2::ggplot(ggplot2::aes(time, p50)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = p05, ymax = p95), alpha = 0.25) +
  ggplot2::geom_line() +
  ggplot2::labs(x = "Time after dose (h)", y = "Daridorexant (ng/mL)") +
  ggplot2::theme_bw()
```

PKNCA on the individual predictions (`Cc`, without residual error):

```{r stoch-pknca}
conc_st <- PKNCA::PKNCAconc(
  dplyr::filter(sim_st, !is.na(Cc)),
  Cc ~ time | treatment + id
)
dose_st <- PKNCA::PKNCAdose(
  dplyr::filter(ev_st, evid == 1) |> dplyr::mutate(treatment = "50 mg bedtime"),
  amt ~ time | treatment + id
)
nca_st <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_st, dose_st, intervals = intervals))
summary(nca_st)

st_res <- as.data.frame(nca_st$result)
med_cmax <- stats::median(st_res$PPORRES[st_res$PPTESTCD == "cmax"])
med_auc <- stats::median(st_res$PPORRES[st_res$PPTESTCD == "auclast"])
c(median_cmax = med_cmax, median_auclast = med_auc)
stopifnot(
  # The cohort median sits near the typical-value reference subject; the
  # cohort's own covariate mix and the log-normal IIV on F shift it by at most
  # tens of percent, while a unit or structural error would shift it by far more.
  abs(log(med_cmax / ref_row$cmax)) < log(1.3),
  abs(log(med_auc / ref_row$auclast)) < log(1.3)
)
```

# Assumptions and deviations

* **Age reference.** The Table 2 footnote gives an age reference of 30 years,
  while the Figure 3 "typical subject" is 60 years old. The model uses the
  footnote's 30 years: with it the reference subject's simulated AUC0-24, Cmax,
  C8h and tmax match Figure 3a and the Results to within 1%, whereas a
  60-year reference lowers AUC0-24 by about 12% and C8h by about 12%.
* **Residual error form.** The paper reports an additive (17.59 ng/mL) and a
  multiplicative (0.16) error term but does not name the Monolix error model.
  The Monolix default combined form `combined1`, `y = f + (a + b*f)*e`, is used.
* **Units of the additive error.** Not printed; ng/mL is used, the unit of the
  paper's concentrations (Km is printed in ug/mL and is converted inside the
  model, where the central concentration is in mg/L).
* **Inter-occasion variability.** Table 2 reports one IOV SD each for F, tlag
  and ka but not how occasions were defined. The model carries four occasion
  slots (`OCC` = 1-4) with equal variances; records with `OCC` outside 1-4 get
  no IOV. The simulations above use `OCC = 1`.
* **Food status encoding.** The paper's three-level food status (fasted, light
  meal / uncontrolled, high-fat high-calorie meal; reference light meal /
  uncontrolled) is carried by the existing `FED` and `FED_HIGHFAT` indicators:
  fasted is `FED = 0`, the reference is `FED = 1, FED_HIGHFAT = 0`, and a
  high-fat meal is `FED = 1, FED_HIGHFAT = 1`.
* **Dosing time.** Bedtime administration is the reference (`DOSETIME_EVENING
  = 1`); morning dosing (`DOSETIME_EVENING = 0`) multiplies ka by exp(1.05).
* **Lean body weight.** The paper's "lean body weight" is computed with the
  Janmahasatian formula, which the covariate register files under `FFM`.
* **Distribution of F.** Table 2 states that all parameters are log-normally
  distributed, so F carries a log-normal IIV and IOV as printed; a sampled
  individual F above 1 is possible for large positive random effects. F is a
  relative bioavailability, so this is not a contradiction in the model.
* **Errata in the Discussion.** The Discussion states "t lag 0.41 and 1.23 h,
  k a 3.8 and 1.05/h in the fasted and fed states". These values do not follow
  from Table 2, whose own footnote derives tlag = 0.25 h (fasted) and 0.76 h
  (fed), matching the Results (15, 25 and 46 min) and the Table 3 ratios
  (tlag fed / fasted = 3.1, ka fasted / fed = 3.6). The model follows Table 2.
  Likewise the Discussion's "1.8 L/h at 1000 ng/mL" does not follow from the
  Table 2 reference Km (which gives 2.07 L/h); it matches a 60-year-old subject
  (Km 2.90 ug/mL). The Table 2 values are used unchanged.
* **Figure 3a lean-body-weight tmax labels.** Figure 3a labels the two
  lean-body-weight tmax end points the opposite way round from what a central
  volume that grows with lean body weight (Table 2, Table 3, and the Cmax panel
  of the same figure) implies. The magnitudes agree with the model; the
  direction is taken from the parameter table, and these two end points are
  left out of the Figure 3 comparison.
* **Virtual cohort.** Weight, height, age and ALP distributions are chosen by
  the maintainers to land near the Table 1 medians; the paper does not give
  joint distributions.
