---
title: "Paracetamol (Krekels 2015)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Paracetamol (Krekels 2015)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Krekels_2015_paracetamol"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Krekels_2015_paracetamol"))$description`
* Article: <https://doi.org/10.1007/s00228-015-1887-y>

The paper develops a one-compartment parent-and-two-metabolite popPK
model for intravenous paracetamol (administered as propacetamol, doses
expressed as paracetamol equivalents) in preterm and term neonates and
infants. Pooled plasma paracetamol concentrations and urinary
recoveries of unchanged paracetamol, paracetamol-glucuronide, and
paracetamol-sulphate were used. The model serves to test whether
glucuronidation upregulates after repeated dosing; the authors conclude
that bodyweight-driven developmental changes alone explain the
observed time trends and that no time-varying (up-regulation) component
is needed.

# Population

Krekels 2015 pools two clinical studies (one single-dose, one
multiple-dose) carried out at University Hospital Leuven (Belgium) and
Erasmus MC-Sophia Children's Hospital (Rotterdam, Netherlands). The
final dataset comprises 54 preterm and term neonates and young infants
with a median bodyweight of 2.5 kg (range 0.5-6.3 kg), median
postnatal age of 1 day (range 1-140 days), and median postmenstrual
age of 36 weeks (range 27-60 weeks). Paracetamol plasma concentrations
were available from all 54 subjects (353 observations); urine
collections in 6, 8, or 12 h aliquots were available for 22 of the 54
subjects (435 observations of unchanged paracetamol or one of the two
metabolites).

The same population information is available programmatically:

```{r population-meta}
local({
  fbody <- body(readModelDb("Krekels_2015_paracetamol"))
  meta_env <- new.env()
  for (stmt in as.list(fbody)[-1]) {
    if (is.call(stmt) && length(stmt) >= 1 &&
        identical(stmt[[1]], as.name("<-"))) {
      eval(stmt, envir = meta_env)
    } else {
      break
    }
  }
  cat("Population:\n")
  str(meta_env$population, max.level = 1)
  cat("\nCovariates:\n")
  str(meta_env$covariateData, max.level = 1)
})
```

# Source trace

Per-parameter and per-equation origin (also recorded as in-file
comments in
`inst/modeldb/specificDrugs/Krekels_2015_paracetamol.R`):

| Equation / parameter         | Value (typical, per kg)                       | Source location          |
|------------------------------|-----------------------------------------------|--------------------------|
| `lvc`                        | `log(1.06)` L/kg                              | Table 1: V1 (RSE 4.34%)  |
| `lcl_gluc`                   | `log(0.266)` mL/min/kg                        | Table 1: CL1 (RSE 17.4%) |
| `lcl_sulf`                   | `log(1.46)` mL/min/kg^n                       | Table 1: CL2 (RSE 14.8%) |
| `lcl_renal`                  | `log(0.285)` mL/min/kg                        | Table 1: CL4 (RSE 6.98%) |
| `e_wt_cl_sulf`               | `1.40`                                        | Table 1: n  (RSE 9.29%)  |
| `lmf`                        | `log(11.3)`                                   | Table 1: mf (RSE 21.5%)  |
| `frac_vmet`                  | `0.18 FIX`                                    | Methods: V2 = V3 = 0.18*V1, fixed from adult data (refs 13-14) |
| `etalvc`                     | `~ 0.0925`                                    | Table 1: IIV V1  (RSE 29.5%) |
| `etalcl_gluc`                | `~ 0.599`                                     | Table 1: IIV CL1 (RSE 41.6%) |
| `etalcl_sulf`                | `~ 0.312`                                     | Table 1: IIV CL2 (RSE 33.3%) |
| `etalcl_renal`               | `~ 0.0879`                                    | Table 1: IIV CL4 (RSE 52.2%) |
| `addSd`                      | `sqrt(0.354)` mg/L                            | Table 1: P plasma additive variance (RSE 29.4%) |
| `propSd`                     | `sqrt(0.0198)`                                | Table 1: P plasma proportional variance (RSE 16.5%) |
| `addSd_urineP`               | `sqrt(0.188)` mg                              | Table 1: P urine additive variance (RSE 27.5%) |
| `addSd_urinePG`              | `sqrt(0.223)` mg                              | Table 1: PG urine additive variance (RSE 14.5%) |
| `addSd_urinePS`              | `sqrt(0.332)` mg                              | Table 1: PS urine additive variance (RSE 35.9%) |
| `d/dt(central)` ODE          | n/a                                           | Figure 1 schematic       |
| `d/dt(central_gluc)` ODE     | n/a                                           | Figure 1 schematic       |
| `d/dt(central_sulf)` ODE     | n/a                                           | Figure 1 schematic       |
| `d/dt(urine)` ODE            | n/a                                           | Figure 1 schematic       |
| `d/dt(urine_gluc)` ODE       | n/a                                           | Figure 1 schematic       |
| `d/dt(urine_sulf)` ODE       | n/a                                           | Figure 1 schematic       |
| Linear bodyweight covariate (Vc, CL_gluc, CL_renal) | n/a                    | Results: linear BW relationships |
| Power bodyweight covariate (CL_sulf)                | `WT^n`, n = 1.40       | Results: exponential equation with estimated n |

# Virtual cohort

The Krekels 2015 simulation in Figure 3 uses a single 'typical
individual' with bodyweight 2.5 kg (the median of the cohort) receiving
a single 15 mg/kg paracetamol-equivalent dose as a 15-minute IV
infusion. We reproduce that scenario directly here and additionally
build a small stochastic VPC across the published bodyweight range
(0.5-6.3 kg) to illustrate inter-individual variability.

```{r cohort}
set.seed(2026L)

dose_mg_per_kg <- 15      # paracetamol-equivalent loading dose
inf_dur_min    <- 15      # 15-minute infusion
horizon_min    <- 36 * 60 # follow-up window (36 h)
times_obs      <- sort(unique(c(seq(0, horizon_min, by = 30),
                                c(6, 12, 18, 24, 30, 36) * 60)))

make_cohort <- function(wt, n = 1L, id_offset = 0L) {
  # wt: vector of body weights (kg); n: number of subjects per weight.
  tibble(
    id = id_offset + seq_len(length(wt) * n),
    WT = rep(wt, each = n)
  ) |>
    mutate(amt_mg = dose_mg_per_kg * WT,
           rate_mg_per_min = amt_mg / inf_dur_min)
}

# Typical 2.5 kg infant (Krekels 2015 Figure 3 'typical individual').
typical <- make_cohort(wt = 2.5, n = 1L)

# Stochastic cohort spanning the published weight range.
vpc_cohort <- make_cohort(wt = round(seq(0.5, 6.3, length.out = 12), 2),
                          n = 5L, id_offset = 10L)

build_events <- function(cohort) {
  doses <- cohort |>
    transmute(id, time = 0, evid = 1L, amt = amt_mg,
              rate = rate_mg_per_min, cmt = "central", WT = WT)
  obs <- cohort |>
    tidyr::expand_grid(time = times_obs) |>
    transmute(id, time, evid = 0L, amt = NA_real_, rate = NA_real_,
              cmt = "Cc", WT = WT)
  bind_rows(doses, obs) |>
    arrange(id, time, desc(evid))
}

events_typ <- build_events(typical)
events_vpc <- build_events(vpc_cohort)

stopifnot(!anyDuplicated(unique(events_typ[, c("id", "time", "evid")])))
stopifnot(!anyDuplicated(unique(events_vpc[, c("id", "time", "evid")])))
```

# Simulation

Two simulations are produced. The first is the Figure 3 'typical
individual' (typical-value parameters via `rxode2::zeroRe()`, 2.5 kg
infant, single 15 mg/kg dose). The second is a stochastic VPC across
the published weight range; we keep IIV non-zero so the spread reflects
the model's between-subject variability for V_c, CL_gluc, CL_sulf, and
CL_renal.

```{r simulate}
mod_full    <- readModelDb("Krekels_2015_paracetamol")
mod_typical <- rxode2::zeroRe(mod_full)

sim_typ <- rxode2::rxSolve(mod_typical, events = events_typ,
                           keep = c("WT")) |>
  as.data.frame() |>
  dplyr::mutate(id = 1L) |>
  dplyr::filter(time >= 0)

sim_vpc <- rxode2::rxSolve(mod_full, events = events_vpc,
                           keep = c("WT")) |>
  as.data.frame() |>
  dplyr::filter(time >= 0)
```

# Replicate Figure 3 (cumulative urinary recovery)

Krekels 2015 Figure 3 plots the relative cumulative amount of
unchanged paracetamol (P), paracetamol-glucuronide (PG), and
paracetamol-sulphate (PS) recovered in urine over 36 hours after a
single 15 mg/kg paracetamol-equivalent dose in a typical 2.5 kg
individual. The recoveries are expressed as molar paracetamol
equivalents normalised to the administered dose (i.e. percent of dose).

```{r figure-3}
dose_typ_mg <- typical$amt_mg

fig3 <- sim_typ |>
  transmute(
    time_h    = time / 60,
    P_pct     = 100 * urineP  / dose_typ_mg,
    PG_pct    = 100 * urinePG / dose_typ_mg,
    PS_pct    = 100 * urinePS / dose_typ_mg
  ) |>
  tidyr::pivot_longer(c(P_pct, PG_pct, PS_pct),
                      names_to = "species", values_to = "pct_dose") |>
  dplyr::mutate(species = factor(species,
                                 levels = c("PS_pct", "PG_pct", "P_pct"),
                                 labels = c("Paracetamol-sulphate (PS)",
                                            "Paracetamol-glucuronide (PG)",
                                            "Unchanged paracetamol (P)")))

ggplot(fig3, aes(time_h, pct_dose, colour = species)) +
  geom_line(linewidth = 0.8) +
  scale_x_continuous(breaks = seq(0, 36, by = 6)) +
  labs(
    x = "Time after dose (h)",
    y = "Cumulative urinary recovery (% of administered dose)",
    colour = NULL,
    title = paste(
      "Replicates Figure 3 of Krekels 2015:",
      "relative cumulative urinary recovery of P, PG, PS"
    ),
    caption = "Typical 2.5 kg individual, single 15 mg/kg paracetamol-equivalent IV infusion over 15 minutes."
  ) +
  theme_minimal()
```

The per-6-h-interval fractions of dose recovered as each species can
be read off the cumulative trajectory by differencing successive
6-hour values. Krekels 2015 Figure 3 annotates these percentages
across the top of the panel; we tabulate them here.

```{r figure-3-interval-table}
interval_table <- sim_typ |>
  dplyr::filter(time %in% (c(6, 12, 18, 24, 30, 36) * 60)) |>
  dplyr::transmute(
    interval_h_end = time / 60,
    P_mg     = urineP,
    PG_mg    = urinePG,
    PS_mg    = urinePS
  ) |>
  dplyr::arrange(interval_h_end) |>
  dplyr::mutate(
    P_in_interval_pct  = 100 * (P_mg  - dplyr::lag(P_mg,  default = 0)) / dose_typ_mg,
    PG_in_interval_pct = 100 * (PG_mg - dplyr::lag(PG_mg, default = 0)) / dose_typ_mg,
    PS_in_interval_pct = 100 * (PS_mg - dplyr::lag(PS_mg, default = 0)) / dose_typ_mg
  ) |>
  dplyr::transmute(
    interval     = paste0(interval_h_end - 6, "-", interval_h_end, " h"),
    P_pct        = round(P_in_interval_pct,  2),
    PG_pct       = round(PG_in_interval_pct, 2),
    PS_pct       = round(PS_in_interval_pct, 2),
    total_pct    = round(P_in_interval_pct + PG_in_interval_pct +
                         PS_in_interval_pct, 2)
  )

knitr::kable(
  interval_table,
  caption = paste(
    "Percentage of administered dose recovered as unchanged paracetamol (P),",
    "paracetamol-glucuronide (PG), and paracetamol-sulphate (PS) in each",
    "6-h urine collection interval, typical 2.5 kg infant."
  )
)
```

The sulphate pathway dominates in this very young population, with
glucuronide the next-largest fraction; unchanged renal excretion is
small. As Krekels 2015 discusses, the glucuronide fraction increases
over time relative to the sulphate fraction not because of
up-regulation of UGT activity but simply because the glucuronide
formation clearance is slower, so the metabolite appears later in
the urine.

# VPC of plasma paracetamol across the published weight range

This figure overlays per-subject plasma paracetamol profiles for
12 weights spanning 0.5-6.3 kg with 5 stochastic subjects per
weight. The 5th, 50th, and 95th simulated-population percentiles are
overlaid as ribbons.

```{r figure-vpc-plasma}
vpc_quantiles <- sim_vpc |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(vpc_quantiles, aes(x = time / 60)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(aes(y = Q50), linewidth = 0.7) +
  scale_y_log10() +
  scale_x_continuous(breaks = seq(0, 36, by = 6)) +
  labs(
    x = "Time after dose (h)",
    y = "Plasma paracetamol Cc (mg/L, log scale)",
    title = "Stochastic VPC of plasma paracetamol (60 subjects spanning 0.5-6.3 kg)",
    caption = "Median (line) and 5th-95th percentiles (ribbon) of simulated plasma Cc."
  ) +
  theme_minimal()
```

# PKNCA validation on plasma paracetamol

PKNCA non-compartmental analysis is run on the typical-value plasma
trajectory (2.5 kg, single 15 mg/kg dose). With only one subject and a
single dose level there is no meaningful between-cohort comparison;
the goal here is to show that Cmax, AUC, and half-life align with the
expected typical-individual values.

```{r pknca}
sim_nca <- sim_typ |>
  dplyr::filter(!is.na(Cc), Cc > 0, time > 0) |>
  dplyr::transmute(id, time_h = time / 60, Cc, cohort = "typical_2.5kg")

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time_h | cohort + id)

dose_df <- events_typ |>
  dplyr::filter(evid == 1L) |>
  dplyr::transmute(id, time_h = time / 60, amt, cohort = "typical_2.5kg")

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time_h | cohort + id)

intervals <- data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- PKNCA::pk.nca(nca_data)

knitr::kable(
  as.data.frame(nca_res$result) |>
    dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
    dplyr::transmute(
      cohort,
      parameter = PPTESTCD,
      value     = signif(PPORRES, 4)
    ) |>
    tidyr::pivot_wider(names_from = parameter, values_from = value),
  caption = paste(
    "Typical-value PKNCA on simulated plasma paracetamol after a single",
    "15 mg/kg paracetamol-equivalent IV (15 min infusion) in a 2.5 kg",
    "infant. Units: Cmax in mg/L, Tmax / half-life in h, AUC[0-Inf] in",
    "mg*h/L."
  )
)
```

## Comparison against published NCA

Krekels 2015 does not report NCA Cmax / AUC / half-life as a separate
results table; its primary outputs are the structural model parameter
estimates (Table 1) and the Figure 3 cumulative urinary recovery
panel. The Cmax and half-life from the PKNCA call above can be
sanity-checked against the structural model: V_c at 2.5 kg is
1.06 * 2.5 = 2.65 L, so a 37.5 mg paracetamol-equivalent dose
delivered as a 15-minute infusion gives an expected end-of-infusion
concentration on the order of 13 mg/L (with infusion-time
distribution lowering the peak slightly from the
instantaneous-bolus 37.5 / 2.65 ~= 14.2 mg/L). The total elimination
rate at 2.5 kg is CL_gluc/V_c + CL_sulf/V_c + CL_renal/V_c =
(0.266 + 1.46 * 2.5^0.40 + 0.285) * 2.5 / 2650 ~= 0.0025 1/min,
implying a half-life of ~4.7 h, consistent with the published
neonatal paracetamol literature.

# Assumptions and deviations

- **Metabolite volume rule.** V_gluc and V_sulf are fixed at 0.18 *
  V_central per Krekels 2015 Methods, citing Anderson et al. (ref 13)
  and Critchley et al. (ref 14). The model file declares
  `frac_vmet <- fixed(0.18)` to preserve this provenance.
- **Shared metabolite urinary-excretion rate constant.** Krekels 2015
  assumes k3 = k5 = mf * k4 with one common multiplication factor
  mf = 11.3 estimated for both glucuronide and sulphate, so the model
  file derives both metabolite urinary-excretion rate constants from
  the single `mf` parameter.
- **Per-kg parameterization.** Table 1 reports each structural
  parameter in per-kg units (L/kg or mL/min/kg, with the sulphate
  formation clearance carrying an estimated power exponent of 1.40).
  The model file's ini() entries therefore represent the typical
  parameter at WT = 1 kg, with linear or power bodyweight scaling
  applied inside model().
- **CL_sulf bodyweight exponent.** Krekels 2015 calls this an
  'exponential' covariate equation; in their notation,
  P_i = P_p * cov^n with n estimated. This is structurally a power
  function (not literally an exponential of bodyweight) and the model
  file implements it as `(WT/wt_ref)^e_wt_cl_sulf`.
- **NONMEM variance -> nlmixr2 SD.** Table 1 reports the variance of
  each residual-error component (and of each IIV eta). The IIV values
  are used directly because the eta is log-normal with omega^2 as the
  internal variance scale; the residual-error values are square-rooted
  in ini() because nlmixr2's `add(...)` and `prop(...)` arguments are
  standard deviations rather than variances.
- **Compartment-naming warnings.** The three urine compartments
  (`urine`, `urine_gluc`, `urine_sulf`) and the observation
  variables (`urineP`, `urinePG`, `urinePS`) follow the same
  cumulative-elimination idiom used by `Pierre_2017_morphine.R`
  (`urine_morphine`, `urine_m3g`) and `Allegaert_2015_paracetamol.R`
  (`urine_apap`, `urine_gluc`, `urine_sulf`). `urine` is not in
  `R/conventions.R`'s canonical compartment list, so
  `checkModelConventions()` emits warnings for these names;
  the convention here is consistent with the existing models for
  the same biological concept (cumulative urinary recovery of
  parent and metabolites).
- **Dose units.** The model expects paracetamol-equivalent mg as the
  dose. Datasets that record propacetamol mg should convert to
  paracetamol equivalents on a molar basis (paracetamol MW 151.16 vs
  propacetamol MW 264.27) before dosing.
- **Observation-time differencing for urine amounts.** Krekels 2015
  fit per-collection-interval urine amounts; the model exposes
  cumulative-urine compartments and observation variables. Downstream
  users who want per-interval recoveries (as in the table replicated
  above) should difference the cumulative output between successive
  collection-interval endpoints.
- **Covariates not retained.** Postnatal age, postmenstrual age, sex,
  term-vs-preterm-birth, and the source-study indicator were tested
  but were not retained in the final Krekels 2015 model; bodyweight
  was the only retained covariate. The model file's `covariateData`
  therefore lists `WT` only.
- **No time-varying (up-regulation) component.** The model file
  deliberately does not include a time-varying scalar on CL_gluc;
  Krekels 2015 specifically tested this and rejected it as
  non-significant. Users investigating time-varying glucuronidation
  hypotheses should add the scalar explicitly rather than expect it
  in this baseline implementation.
