---
title: "Clonazepam (Kruizinga 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clonazepam (Kruizinga 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Kruizinga_2022_clonazepam"))$reference`
* Description: `r rxode2::rxode(readModelDb("Kruizinga_2022_clonazepam"))$description`
* Article: <https://doi.org/10.1111/bcp.15152> (open access; PMC9299763)

Kruizinga 2022 fits a joint plasma + saliva population PK model for
clonazepam in healthy young adults, as a step towards a clinical trial in
children with ARID1B-related intellectual disability in which blood sampling
would be too burdensome. The model is then used to show by simulation that a
plasma trough concentration can be predicted from one to five saliva samples
by Bayesian MAP estimation.

# Population

Twenty healthy subjects aged 18-30 years (mean 22.4, SD 2.8) took part at the
Centre for Human Drug Research, Leiden, the Netherlands, in June-July 2020
(Kruizinga 2022 Methods 2.1-2.2 and Table 1). Nine were male (45%), all were
Caucasian, and mean (SD) weight was 67.8 (8.3) kg, height 175.1 (7.3) cm and
BMI 22.2 (2.4) kg/m^2. Each subject took a single oral dose of 0.5 mg (n = 10)
or 1.0 mg (n = 10) clonazepam solution (Rivotril drops) in lemonade. Paired
plasma and saliva samples were taken at 0.5, 1, 2, 4, 6, 8, 24 and 48 h; mouths
were rinsed with water 10 minutes before each saliva sample, which was
collected with a SalivaBio Infant Swab. All 160 plasma samples and 154 of the
160 saliva samples were analysable, and none was below the limit of
quantification.

The same information is available programmatically from the model's
`population` metadata:

```{r population}
pop <- rxode2::rxode(readModelDb("Kruizinga_2022_clonazepam"))$population
str(pop, max.level = 1)
```

# Model structure

**Plasma.** Two-compartment disposition with first-order absorption from a
depot. Allometric scaling is fixed and referenced to 70 kg: exponent 0.75 on
clearance and intercompartmental clearance, 1 on both volumes. The relative
bioavailability is fixed to 1 with between-subject variability (F plasma).
The between-subject variance on ka was 0.66 and bimodal, so the authors added a
two-class **mixture on ka**. The fast class has ka fixed at 100 1/h and no
variability. The slow class (estimated probability 0.75) has ka = 1.106 1/h with
variability. In the packaged model the class is the binary covariate
`MIX_FAST_ABS` (1 = fast class), drawn per subject as Bernoulli(0.25) for a
population simulation.

**Saliva.** The oral solution leaves a residue in the mouth that dominates the
saliva concentration for the first ~4 h. The paper models it with a 1 mL
*contamination compartment*. It receives a fraction `F saliva` of the dose,
0.033 per mille, and empties first-order (Equation 1). After that the saliva
concentration is the plasma concentration times a saliva:plasma ratio that
rises with plasma concentration towards a ceiling (Equations 2-3):

$$\text{ratio} = \text{Ratio}_{MAX} \cdot \frac{C_p}{C_p + \text{Ratio}_{KM}}, \qquad
C_{saliva} = \frac{\text{Contamination}}{0.001\,\text{L}} + C_p \cdot \text{ratio}$$

The packaged model uses the canonical `saliva` compartment for the
contamination term, with `fcontam_saliva`, `kel_saliva` and the fixed volume
`vsaliva` = 0.001 L. The saturable ratio is `fsaliva`, computed from
`fsaliva_max` and `km_fsaliva`.

**Dosing the contamination compartment.** The contamination compartment is not
connected to `central`, so the residue is put in by dosing it directly. Every
dose therefore appears **twice** in the event table: once to `depot` and once,
with the same amount, to `saliva`. The model applies `f(saliva) <- fcontam_saliva`,
so the saliva record delivers only the residue fraction. The residue is a
measurement artefact: it never reaches `central` and does not reduce the
systemic dose.

# Source trace

Every `ini()` value carries an in-file comment pointing to its source. The
table below collects them.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` (slow class) | log(1.106) 1/h | Table 2, 'ka - slow group' |
| `lka_fastabs` (fast class) | fixed(log(100)) 1/h | Results 3.1 ('fixed ka of 100/h') |
| `MIX_FAST_ABS` probability | 0.25 (= 1 - 0.75) | Table 2, 'Prob. slow group' = 0.75 |
| `lcl` | log(2.98) L/h | Table 2, 'clearance' |
| `lvc` | log(109.5) L | Table 2, 'VD central' |
| `lq` | log(61.37) L/h | Table 2, 'Q' |
| `lvp` | log(130.6) L | Table 2, 'VD peripheral' |
| `lfdepot` | fixed(log(1)) | Methods 2.4 (no bioavailability estimable from oral-only data) |
| `e_wt_cl` (CL, Q) | fixed(0.75) | Methods 2.4; Table 2 footnote |
| `e_wt_vc` (Vc, Vp) | fixed(1) | Methods 2.4; Table 2 footnote |
| `lfcontam_saliva` | log(0.033e-3) | Table 2, 'F saliva' = 0.033 (% 1000^-1, per mille) |
| `lkel_saliva` | log(1.95) 1/h | Table 2, 'Kel saliva'; Equation 1 |
| `lvsaliva` | fixed(log(0.001)) L | Results 3.1 (1 mL fixed); Equation 3 |
| `lfsaliva_max` | log(0.195) | Table 2, 'RatioMAX'; Equation 2 |
| `lkm_fsaliva` | log(2.581) ug/L | Table 2, 'RatioKM'; Equation 2 |
| `etalka` | 0.16 | Table 2, 'omega2 ka - slow group' |
| `etalq` | 0.25 | Table 2, 'omega2 Q' |
| `etalfdepot` | 0.026 | Table 2, 'omega2 F plasma' |
| `etalfcontam_saliva` | 0.28 | Table 2, 'omega2 F saliva' |
| `etalkel_saliva` | 0.056 | Table 2, 'omega2 Kel saliva' |
| `propSd` (plasma) | sqrt(0.0058) = 0.0762 | Table 2, 'sigma2 proportional plasma' (variance) |
| `propSd_Csaliva` | sqrt(0.057) = 0.239 | Table 2, 'sigma2 proportional saliva' (variance) |
| `d/dt(saliva)` | n/a | Equation 1 |
| `fsaliva` | n/a | Equation 2; Table 2 footnote 'TVRatio' |
| `Csaliva` | n/a | Equation 3 |

The Table 2 note states that omega2 and sigma2 are variances, so the residual
standard deviations are their square roots.

# Typical-value check against Figure 1

Figure 1 of the paper shows the observed mean (SD) plasma and saliva profiles
for each dose group. The maintainers digitised the plotted means from the
published figure; the 1 mg saliva panel has no plotted mean at 0.5 h. A
typical subject (67.8 kg, the Table 1 mean weight; slow absorption class, the
majority phenotype) is solved for each dose.

```{r fig1-digitised}
fig1_obs <- tibble::tribble(
  ~dose_mg, ~time, ~plasma, ~saliva,
  0.5, 0.5, 2.42, 9.70,
  0.5, 1,   2.62, 2.73,
  0.5, 2,   2.69, 0.713,
  0.5, 4,   2.46, 0.303,
  0.5, 6,   2.27, 0.230,
  0.5, 8,   2.14, 0.206,
  0.5, 24,  1.92, 0.138,
  0.5, 48,  1.41, 0.091,
  1.0, 0.5, 4.22, NA,
  1.0, 1,   4.65, 4.20,
  1.0, 2,   4.71, 1.28,
  1.0, 4,   3.80, 0.538,
  1.0, 6,   3.56, 0.391,
  1.0, 8,   3.37, 0.365,
  1.0, 24,  2.78, 0.217,
  1.0, 48,  1.98, 0.158
)
```

```{r typical}
mod <- readModelDb("Kruizinga_2022_clonazepam")
mod_typ <- rxode2::zeroRe(mod)

make_single_dose <- function(n, dose_mg, times, id_offset = 0L) {
  subj <- tibble(id = id_offset + seq_len(n), dose_mg = dose_mg)
  doses <- subj |>
    tidyr::crossing(cmt = c("depot", "saliva")) |>
    mutate(time = 0, amt = dose_mg, evid = 1L, dvid = NA_integer_)
  obs <- subj |>
    tidyr::crossing(time = times) |>
    mutate(cmt = "central", amt = NA_real_, evid = 0L, dvid = 1L)
  bind_rows(doses, obs) |>
    arrange(id, time, desc(evid))
}

fig1_times <- c(0.5, 1, 2, 4, 6, 8, 24, 48)
ev_typ <- bind_rows(
  make_single_dose(1, 0.5, fig1_times, id_offset = 0L),
  make_single_dose(1, 1.0, fig1_times, id_offset = 1L)
) |>
  mutate(WT = 67.8, MIX_FAST_ABS = 0L)

sim_typ <- rxode2::rxSolve(
  mod_typ,
  events = ev_typ,
  keep = "dose_mg",
  useLinCmt = FALSE # rxode2's ODE->linCmt conversion breaks the dvid mapping
) |>
  as.data.frame()

cmp_fig1 <- sim_typ |>
  select(dose_mg, time, Cc, Csaliva) |>
  inner_join(fig1_obs, by = c("dose_mg", "time")) |>
  mutate(
    ratio_plasma = Cc / plasma,
    ratio_saliva = Csaliva / saliva
  )

cmp_fig1 |>
  mutate(across(c(Cc, Csaliva, ratio_plasma, ratio_saliva), ~ signif(.x, 3))) |>
  rename(
    "Dose (mg)" = dose_mg,
    "Time (h)" = time,
    "Plasma, typical (ug/L)" = Cc,
    "Plasma, Fig 1 mean (ug/L)" = plasma,
    "Plasma ratio" = ratio_plasma,
    "Saliva, typical (ug/L)" = Csaliva,
    "Saliva, Fig 1 mean (ug/L)" = saliva,
    "Saliva ratio" = ratio_saliva
  ) |>
  knitr::kable(caption = "Typical-value prediction vs digitised Figure 1 means.")
```

The observed data are not dose-proportional. The 48 h dose-normalised plasma
mean is 2.8 ug/L per mg in the 0.5 mg group and 2.0 ug/L per mg in the 1 mg
group. A linear model fitted to both groups therefore runs about 15% below the
0.5 mg means and about 15% above the 1 mg means. The same split carries
through to saliva, amplified by the saturable ratio. Pooled across the two
groups, the post-absorption phase agrees closely. The gate below checks that
pooled agreement. Plasma is checked from 4 h. Saliva is checked from 6 h,
because before then the contamination term dominates.

```{r typical-gate}
late_plasma <- cmp_fig1 |> filter(time >= 4)
late_saliva <- cmp_fig1 |> filter(time >= 6, !is.na(saliva))
gm <- function(x) exp(mean(log(x)))
gm_plasma <- gm(late_plasma$ratio_plasma)
gm_saliva <- gm(late_saliva$ratio_saliva)
c(gm_plasma = gm_plasma, gm_saliva = gm_saliva)

# Deterministic (typical-value) solve against digitised means: a mis-scaled
# CL, V, dose or unit moves every ratio together and breaks the pooled
# geometric mean; the per-point envelope allows for the dose-group imbalance
# described above (pooled plasma GM 0.98, per-point 0.82-1.18; saliva GM
# 1.02, per-point 0.73-1.55 at the time of writing).
stopifnot(
  abs(log(gm_plasma)) < log(1.15),
  all(late_plasma$ratio_plasma > 0.7 & late_plasma$ratio_plasma < 1.4),
  abs(log(gm_saliva)) < log(1.25),
  all(late_saliva$ratio_saliva > 0.5 & late_saliva$ratio_saliva < 2)
)
```

# Virtual cohort for the study design

A stochastic version of the study design: 100 virtual subjects per dose group.
Weight is drawn from a normal distribution matching Table 1 (67.8, SD 8.3 kg).
Absorption class is drawn as Bernoulli(0.25) for the fast class.

```{r cohort}
rxode2::rxSetSeed(20220601)
set.seed(20220601)

n_arm <- 100
fine_times <- sort(unique(c(seq(0.25, 8, by = 0.25), seq(9, 48, by = 1))))
ev_study <- bind_rows(
  make_single_dose(n_arm, 0.5, fine_times, id_offset = 0L),
  make_single_dose(n_arm, 1.0, fine_times, id_offset = n_arm)
)
stopifnot(!anyDuplicated(unique(ev_study[, c("id", "time", "evid", "cmt")])))

subj_cov <- tibble(
  id = seq_len(2 * n_arm),
  WT = pmin(pmax(rnorm(2 * n_arm, 67.8, 8.3), 45), 95),
  MIX_FAST_ABS = rbinom(2 * n_arm, 1, 0.25)
)
ev_study <- ev_study |>
  left_join(subj_cov, by = "id") |>
  mutate(treatment = paste(dose_mg, "mg"))

sim_study <- rxode2::rxSolve(
  mod,
  events = ev_study,
  keep = c("treatment", "dose_mg", "WT", "MIX_FAST_ABS"),
  useLinCmt = FALSE
) |>
  as.data.frame()
```

## Figure 1 -- simulated profiles with the observed means

```{r figure-1, fig.width = 8, fig.height = 6}
vpc <- sim_study |>
  select(treatment, time, Plasma = Cc, Saliva = Csaliva) |>
  pivot_longer(c(Plasma, Saliva), names_to = "matrix", values_to = "conc") |>
  group_by(treatment, matrix, time) |>
  summarise(
    Q10 = quantile(conc, 0.10),
    Q50 = median(conc),
    Q90 = quantile(conc, 0.90),
    .groups = "drop"
  )

obs_long <- fig1_obs |>
  mutate(treatment = paste(dose_mg, "mg")) |>
  select(treatment, time, Plasma = plasma, Saliva = saliva) |>
  pivot_longer(c(Plasma, Saliva), names_to = "matrix", values_to = "conc") |>
  filter(!is.na(conc))

ggplot(vpc, aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue4") +
  geom_point(data = obs_long, aes(time, conc), inherit.aes = FALSE) +
  facet_grid(matrix ~ treatment, scales = "free_y") +
  scale_y_log10() +
  labs(
    x = "Time (h)",
    y = "Clonazepam concentration (ug/L)",
    title = "Simulated median and 80% interval vs observed means",
    caption = "Replicates Figure 1 of Kruizinga 2022 (points: digitised observed means)."
  )
```

The early saliva peak is the contamination term. It falls with the 1.95 1/h
`kel_saliva` (half-life about 21 minutes). From about 4 h on, saliva tracks
plasma through the saturable ratio, as the paper describes.

## PKNCA on the study design

The paper reports no NCA, so this table gives the simulated plasma
exposures for reference only. The comparison against published values is the
trough check in the next section.

```{r pknca-study}
nca_conc <- sim_study |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)
nca_conc <- bind_rows(
  nca_conc,
  nca_conc |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

nca_dose <- ev_study |>
  filter(evid == 1, cmt == "depot") |>
  select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(
  nca_conc, Cc ~ time | treatment + id,
  concu = "ug/L", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(nca_dose, amt ~ time | treatment + id, doseu = "mg")
intervals <- data.frame(
  start = 0, end = 48,
  cmax = TRUE, tmax = TRUE, auclast = TRUE
)
nca_study <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

as.data.frame(nca_study$result) |>
  group_by(treatment, PPTESTCD) |>
  summarise(median = signif(median(PPORRES), 3), .groups = "drop") |>
  pivot_wider(names_from = PPTESTCD, values_from = median) |>
  rename(
    "Dose group" = treatment,
    "Cmax (ug/L)" = cmax,
    "Tmax (h)" = tmax,
    "AUC0-48 (h*ug/L)" = auclast
  ) |>
  knitr::kable(caption = "Simulated plasma NCA (median) for the study design.")
```

# Replicating the paper's simulation (Figure 2A and Table 3)

Methods 2.5 describes the paper's simulation cohort. Age is uniform between 6
and 30 years, with weights between the 10th and 90th centiles for age. Dosing
is 0.015 mg/kg twice daily, capped at 0.5 mg per dose. The F plasma variability
is raised to a 50% CV, because the healthy-volunteer estimate was thought too
narrow for patients. The paper reports a median plasma trough of
**2.1 ug/L after dose 1** and **13.7 ug/L at steady state** (240 h, after dose
20; Table 3 header). Figure 2A shows the median and 80% prediction interval in
plasma and saliva over the 240 h.

The paper does not tabulate the centile curves it used. The cohort below
approximates them with a sex-averaged median weight-for-age curve (WHO/CDC
style). Each subject's median is multiplied by a uniform factor between 0.82
and 1.25, an approximate 10th-90th centile band. The cohort is capped at 200
subjects; the paper simulated 2000.

```{r cohort-paper-sim}
rxode2::rxSetSeed(20220602)
set.seed(20220602)

n_sim <- 200
wt_ref_age <- c(6, 8, 10, 12, 14, 16, 18, 30)
wt_ref_kg <- c(20.5, 25.3, 31.9, 40.0, 49.0, 57.0, 62.0, 72.0)

paper_cohort <- tibble(
  id = seq_len(n_sim),
  AGE = runif(n_sim, 6, 30),
  WT = approx(wt_ref_age, wt_ref_kg, xout = AGE)$y * runif(n_sim, 0.82, 1.25),
  MIX_FAST_ABS = rbinom(n_sim, 1, 0.25)
) |>
  mutate(dose_mg = pmin(0.015 * WT, 0.5))

dose_times <- seq(0, 228, by = 12) # 20 doses; the steady-state trough is at 240 h
obs_times <- seq(0.5, 240, by = 0.5)

ev_paper <- bind_rows(
  paper_cohort |>
    tidyr::crossing(time = dose_times, cmt = c("depot", "saliva")) |>
    mutate(amt = dose_mg, evid = 1L, dvid = NA_integer_),
  paper_cohort |>
    tidyr::crossing(time = obs_times) |>
    mutate(cmt = "central", amt = NA_real_, evid = 0L, dvid = 1L)
) |>
  arrange(id, time, desc(evid)) |>
  mutate(treatment = "0.015 mg/kg BID")

# F plasma CV raised to 50% (Methods 2.5): log-normal variance log(1 + 0.5^2).
mod_paper_sim <- mod |> rxode2::ini(etalfdepot ~ 0.2231436)

sim_paper <- rxode2::rxSolve(
  mod_paper_sim,
  events = ev_paper,
  keep = c("treatment", "WT", "AGE"),
  useLinCmt = FALSE
) |>
  as.data.frame()
```

```{r figure-2a, fig.width = 8, fig.height = 4.5}
sim_paper |>
  select(time, Plasma = Cc, Saliva = Csaliva) |>
  pivot_longer(c(Plasma, Saliva), names_to = "matrix", values_to = "conc") |>
  group_by(matrix, time) |>
  summarise(
    Q10 = quantile(conc, 0.10),
    Q50 = median(conc),
    Q90 = quantile(conc, 0.90),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50, colour = matrix, fill = matrix)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.2, colour = NA) +
  geom_line() +
  scale_y_log10() +
  scale_colour_manual(values = c(Plasma = "black", Saliva = "blue")) +
  scale_fill_manual(values = c(Plasma = "grey40", Saliva = "blue")) +
  labs(
    x = "Time after first dose (h)",
    y = "Concentration (ug/L)",
    colour = NULL, fill = NULL,
    title = "Median and 80% prediction interval, 0.015 mg/kg BID",
    caption = "Replicates Figure 2A of Kruizinga 2022."
  )
```

## Comparison against published trough medians

PKNCA computes the trough as the last observation in each 12 h interval
(`clast.obs`): after dose 1 over 0-12 h, and at steady state over 228-240 h.
The trough uses `Cc`, the individual prediction without residual error. That
matches the paper's 'true' trough, which comes from the simulated individual
parameters.

```{r pknca-trough}
trough_conc <- sim_paper |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)
trough_conc <- bind_rows(
  trough_conc,
  trough_conc |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

trough_dose <- ev_paper |>
  filter(evid == 1, cmt == "depot") |>
  select(id, time, amt, treatment)

trough_intervals <- data.frame(
  start = c(0, 228), end = c(12, 240),
  clast.obs = TRUE
)
nca_trough <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(trough_conc, Cc ~ time | treatment + id, concu = "ug/L", timeu = "h"),
  PKNCA::PKNCAdose(trough_dose, amt ~ time | treatment + id, doseu = "mg"),
  intervals = trough_intervals
))

sim_trough <- as.data.frame(nca_trough$result) |>
  mutate(window = ifelse(start == 0, "After dose 1 (12 h)", "Steady state (240 h)")) |>
  select(window, PPTESTCD, PPORRES)

published_trough <- tibble::tribble(
  ~window, ~clast.obs,
  "After dose 1 (12 h)", 2.1,
  "Steady state (240 h)", 13.7
)

cmp_trough <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_trough,
  reference = published_trough,
  by = "window",
  units = c(clast.obs = "ug/L"),
  tolerance_pct = 20
)
knitr::kable(
  cmp_trough,
  caption = "Median plasma trough: simulated vs Kruizinga 2022 Table 3 header."
)
```

Both medians fall within the 20% tolerance. Most of the remaining difference
is Monte-Carlo noise in a 200-subject cohort plus the approximate
weight-for-age curve. Across repeated cohorts the simulated medians centre
within a few percent of the published 2.1 and 13.7 ug/L.

```{r trough-gate}
trough_med <- sim_trough |>
  group_by(window) |>
  summarise(median = median(PPORRES), .groups = "drop") |>
  left_join(published_trough, by = "window") |>
  mutate(pct_diff = 100 * (median - clast.obs) / clast.obs)
trough_med

# The median of 200 subjects moves with the drawn cohort: over seven seeds it
# ranged from -10% to +11% of the published values (centred near +3%), and the
# seed used here sits at the low end. A mis-scaled CL, volume, F or dose moves
# both medians by tens of percent.
stopifnot(all(abs(trough_med$pct_diff) < 20))
```

# Assumptions and deviations

- **Figure 1 means are digitised.** Kruizinga 2022 does not tabulate the
  observed concentrations. The maintainers digitised the plotted means from
  Figure 1 by pixel position against the axis ticks. The 1 mg saliva panel
  has no plotted mean at 0.5 h.
- **Absorption mixture as a covariate.** The NONMEM mixture is encoded as
  the binary covariate `MIX_FAST_ABS`, which the user supplies (1 = fast class,
  ka fixed at 100 1/h). For a population simulation draw it as
  Bernoulli(0.25); for a typical-value simulation of the majority phenotype set
  it to 0. Only the slow class carries `etalka`, matching Table 2 ('omega2 ka -
  slow group').
- **Each dose is entered twice.** The contamination compartment receives the
  dose directly, so each dose needs a `depot` record and a `saliva` record of
  the same amount; `f(saliva)` applies the 0.033 per-mille residue fraction.
  Leaving out the `saliva` record removes the early contamination peak but
  leaves the plasma-driven saliva term (and all plasma predictions) unchanged.
- **Unit of F saliva.** Table 2 prints `F saliva` as 0.033 in units of
  '% 1000^-1', which the maintainers read as per mille (3.3e-5 of the dose).
  With a percent reading (3.3e-4), the 0.5 h saliva prediction would be about
  64 ug/L. The plotted means at 0.5 h are below 10 ug/L.
- **Unit scaling.** Dose is in mg and volumes in L, so `Cc` and `Csaliva` are
  multiplied by 1000 to give ug/L, the units of the paper's figures and of
  `RatioKM`.
- **Simulation cohort.** The paper's weight-for-age centile curves are not
  given. They are approximated by a sex-averaged median weight-for-age curve
  times a uniform 0.82-1.25 factor. The cohort has 200 subjects (the paper used
  2000). The F plasma variance for this simulation is `log(1 + 0.5^2)` =
  0.223, the log-normal variance for a 50% CV. The paper does not say whether
  it used this or `0.5^2`.
- **Albumin** was tested as a covariate on the saliva:plasma relationship and
  not retained (Discussion). It is listed in `covariatesDataExcluded`.
- **Errata.** EuropePMC lists no correction notice linked to Kruizinga 2022
  (PMID 34811788) as of 2026-10-08.
