---
title: "Naldemedine (Kubota 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Naldemedine (Kubota 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Kubota_2018_naldemedine"))$reference`
* Article: <https://doi.org/10.1007/s11095-018-2501-7> (PMCID: PMC6182381, open access)
* Supplementary material (Supplemental Tables S1-S7 and Figures S1-S6):
  <https://doi.org/10.1007/s11095-018-2501-7>

Naldemedine is a peripherally acting mu-opioid receptor antagonist for
opioid-induced constipation (OIC). Kubota 2018 pooled 8,146 plasma
concentrations from 949 subjects across 18 phase 1 to phase 3 studies -- healthy
volunteers, patients with chronic non-cancer pain and OIC, and cancer patients
with OIC -- into a single population PK model, then used empirical Bayes
steady-state exposures from that model to fit logistic exposure-response models
for efficacy and for gastrointestinal safety.

This vignette covers **eight** models from that one paper:

* `Kubota_2018_naldemedine` -- the population PK model (two compartments,
  first-order absorption, absorption lag time).
* Seven landmark exposure-response logistic regressions, packaged as the
  `Kubota_2018_naldemedine_*` family and validated in the *Exposure-response
  models* section below, which also closes the loop by feeding PK-simulated
  steady-state exposures into them.

The seven are split the way the authors fitted them. Kubota 2018 analysed the
phase 2b study and the two pooled pivotal phase 3 studies **separately**,
"since the study duration and definition of SBM responders were different", and
fitted each gastrointestinal severity threshold as its own logistic regression.
Replicating that structure means seven independent two-parameter fits, not one
ordinal model.

```{r model}
mod <- rxode2::rxode(readModelDb("Kubota_2018_naldemedine"))
mod
```

# Population

```{r population}
pop <- mod$population
tibble::tibble(
  Field = names(pop),
  Value = vapply(
    pop,
    function(x) paste(if (is.null(names(x))) as.character(x) else paste0(names(x), ": ", x), collapse = "; "),
    character(1)
  )
) |>
  knitr::kable(caption = "Population metadata (Kubota 2018 Tables I and II).")
```

The reference subject against which every covariate effect is centred is, in the
paper's own words, "52-year-old, 76 kg male, white, non-cancer, OIC patients
with CLcr of 108 ml/min, administed under fasted condintion". Note that this is
a **patient with chronic non-cancer pain**, not a healthy volunteer -- which is
why the reference Vc/F of 91.1 L is `THETA(6) * 1.20`, not `THETA(6)`.

# Source trace

Every structural value in the model file, with its source location.

```{r source-trace-pk}
tibble::tribble(
  ~Quantity,                  ~`Model parameter`, ~Value,     ~`Source location`,
  "CL/F, reference",          "lcl",              "9.10 L/h", "Table III THETA(1)",
  "Age exponent on CL/F",     "e_age_cl",         "-0.195",   "Table III THETA(2); Eq. (1)",
  "CLcr exponent on CL/F",    "e_crcl_cl",        "0.0739",   "Table III THETA(3); Eq. (1)",
  "Non-White factor on CL/F", "e_nonwhite_cl",    "0.870",    "Table III THETA(4); Eq. (2)",
  "Female factor on CL/F",    "e_female_cl",      "0.902",    "Table III THETA(5); Eq. (2)",
  "Vc/F, healthy at 76 kg",   "lvc",              "75.9 L",   "Table III THETA(6)",
  "Weight exponent on Vc/F",  "e_wt_vc",          "1 (fixed)", "Table III writes '* (Body weight/76)' with no exponent",
  "Non-cancer factor Vc/F",   "e_cncp_vc",        "1.20",     "Table III THETA(7); Eq. (2)",
  "Cancer factor on Vc/F",    "e_cancer_vc",      "1.27",     "Table III THETA(8); Eq. (2)",
  "Fed factor on Vc/F",       "e_fed_vc",         "1.12",     "Table III THETA(9); Eq. (2)",
  "Ka, reference",            "lka",              "2.94 1/h", "Table III THETA(10)",
  "Age exponent on Ka",       "e_age_ka",         "-1.16",    "Table III THETA(11); Eq. (1)",
  "Q/F",                      "lq",               "4.77 L/h", "Table III",
  "Vp/F",                     "lvp",              "41.8 L",   "Table III",
  "ALAG",                     "ltlag",            "0.195 h",  "Table III",
  "IIV CL/F",                 "etalcl",           "37.9% CV", "Table III",
  "IIV Vc/F",                 "etalvc",           "25.3% CV", "Table III",
  "IIV Ka",                   "etalka",           "161.2% CV", "Table III",
  "IIV Q/F",                  "etalq",            "46.3% CV", "Table III",
  "IIV Vp/F",                 "etalvp",           "36.3% CV", "Table III",
  "Proportional residual",    "propSd",           "25.7% CV", "Table III, intra-individual variability"
) |>
  knitr::kable(caption = "Source trace for the population PK model.")
```

The exposure-response coefficients are traced in their own section below.

## Which CV convention does Table III use?

Table III reports inter-individual variability as a percent CV without saying
whether it is the approximate CV (`100 * sqrt(omega^2)`) or the exact
log-normal CV (`100 * sqrt(exp(omega^2) - 1)`). The two differ enormously for
Ka, whose CV is 161.2%: `omega^2` is 2.599 under the first reading and 1.281
under the second.

The printed 95% confidence intervals settle it. NONMEM's covariance step
produces an interval that is **symmetric on the estimated variance**, so
back-transforming the printed CV endpoints with the correct formula must
restore that symmetry. The test below does exactly that, and is decisive on the
two widest intervals -- which are also the two where the readings diverge most.

```{r omega-scale}
cvtab <- tibble::tribble(
  ~Parameter, ~CV,    ~Lower, ~Upper,
  "CL/F",      37.9,   35.2,   40.5,
  "Vc/F",      25.3,   20.8,   29.2,
  "Ka",       161.2,  142.7,  177.9,
  "Q/F",       46.3,   29.9,   58.2,
  "Vp/F",      36.3,   30.2,   41.6
) |>
  dplyr::mutate(
    # Half-width ratio (upper / lower) of the back-transformed interval.
    # A value near 1 means the transform restored the symmetry NONMEM produced.
    approx_ratio = ((Upper / 100)^2 - (CV / 100)^2) / ((CV / 100)^2 - (Lower / 100)^2),
    exact_ratio = (log(1 + (Upper / 100)^2) - log(1 + (CV / 100)^2)) /
      (log(1 + (CV / 100)^2) - log(1 + (Lower / 100)^2)),
    `omega^2 used` = (CV / 100)^2
  )

cvtab |>
  dplyr::select(
    Parameter, `CV%` = CV, Lower, Upper,
    `Symmetry, omega2=(CV)^2` = approx_ratio,
    `Symmetry, omega2=log(1+CV2)` = exact_ratio,
    `omega^2 used`
  ) |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 4))) |>
  knitr::kable(caption = "Back-transformed CI symmetry under the two readings of Table III's CV% column. A ratio near 1 is the correct reading.")

# Ka and Q/F have by far the widest intervals and therefore the most
# discriminating power; the other three are near-linear under both transforms
# and cannot decide. These bounds are deterministic arithmetic on printed
# numbers -- no simulation is involved -- so they are gated tightly.
disc <- cvtab |> dplyr::filter(Parameter %in% c("Ka", "Q/F"))
stopifnot(
  nrow(disc) == 2L,
  max(abs(disc$approx_ratio - 1)) < 0.02,
  min(abs(disc$exact_ratio - 1)) > 0.08
)

# And the values actually packaged in the model file match that reading.
packaged <- mod$iniDf$est[match(c("etalcl", "etalvc", "etalka", "etalq", "etalvp"), mod$iniDf$name)]
stopifnot(max(abs(packaged - cvtab$`omega^2 used`)) < 1e-6)
```

The approximate reading is symmetric to within 0.7% on Ka and 0.5% on Q/F; the
exact log-normal reading is 14% and 10% out. Table III's CV% column is
`100 * sqrt(omega^2)`, and that is what the model file carries.

# Structural checks (typical value, no random effects)

## The reference subject reproduces the published anchors

```{r reference-subject}
ref_cov <- data.frame(
  AGE = 52, WT = 76, CRCL = 108, SEXF = 0, RACE_WHITE = 1,
  DIS_HEALTHY = 0, DIS_CANCER = 0, FED = 0
)

solve_typical <- function(ui, cov, ev) {
  ev <- as.data.frame(ev)
  ev <- cbind(ev, cov[rep(1L, nrow(ev)), , drop = FALSE])
  rxode2::rxSolve(rxode2::zeroRe(ui), ev, returnType = "data.frame")
}

ev_single <- rxode2::et(amt = 0.2, cmt = "depot") |>
  rxode2::et(seq(0, 72, by = 0.01), cmt = "central")

ref <- solve_typical(mod, ref_cov, ev_single)
ref <- ref[!is.na(ref$Cc), ]

anchors <- tibble::tribble(
  ~Quantity,   ~Published, ~Simulated,
  "CL/F (L/h)",    9.10,   ref$cl[1],
  "Vc/F (L)",     91.1,    ref$vc[1],
  "Ka (1/h)",      2.94,   ref$ka[1],
  "Q/F (L/h)",     4.77,   ref$q[1],
  "Vp/F (L)",     41.8,    ref$vp[1],
  "ALAG (h)",      0.195,  ref$tlag[1]
) |>
  dplyr::mutate(`% difference` = round(100 * (Simulated - Published) / Published, 4))

anchors |>
  dplyr::mutate(Simulated = round(Simulated, 4)) |>
  knitr::kable(caption = "Reference-subject parameters against Kubota 2018 Results. Vc/F 91.1 L is THETA(6) x 1.20, because the reference subject is a chronic-non-cancer-pain patient, not a healthy volunteer.")

# Deterministic: these are algebra on the published THETAs with the random
# effects zeroed, so the only tolerance needed is the printing precision of the
# paper's own 3-significant-figure values.
stopifnot(max(abs(anchors$`% difference`)) < 0.05)
```

## The ODE system really is two-compartment

A `cl` / `vc` pair in `model()` can make rxode2 substitute an analytic solution
for the explicit `d/dt` block. That substitution is only harmless when the
parameter set is complete: with `q` and `vp` missing, a two-compartment model
silently collapses to one compartment and every AUC-based check still passes.
This model does supply `q` and `vp`, so both readings should coincide -- but
"should" is not a check. The gate below compares the solve against an
independently written closed-form two-compartment first-order-absorption
solution with lag.

```{r closed-form}
p <- ref[1, c("cl", "vc", "ka", "q", "vp", "tlag")]
kel <- p$cl / p$vc
k12 <- p$q / p$vc
k21 <- p$q / p$vp
bsum <- kel + k12 + k21
disc2 <- sqrt(bsum^2 - 4 * kel * k21)
alpha <- (bsum + disc2) / 2
beta <- (bsum - disc2) / 2

tt <- pmax(ref$time - p$tlag, 0)
cf <- 1000 * 0.2 * p$ka / p$vc * (
  (k21 - p$ka) / ((alpha - p$ka) * (beta - p$ka)) * exp(-p$ka * tt) +
    (k21 - alpha) / ((p$ka - alpha) * (beta - alpha)) * exp(-alpha * tt) +
    (k21 - beta) / ((p$ka - beta) * (alpha - beta)) * exp(-beta * tt)
)
cf[ref$time < p$tlag] <- 0

keep <- cf > 1e-6
rel_err <- max(abs(ref$Cc[keep] - cf[keep]) / cf[keep])

cat(sprintf(
  "alpha = %.4f 1/h, beta = %.4f 1/h (t1/2 = %.2f h); max relative error = %.2e over %d points\n",
  alpha, beta, log(2) / beta, rel_err, sum(keep)
))

stopifnot(
  # Numerical solver agreement, not a cohort statistic: tight.
  rel_err < 1e-4,
  # Two DISTINCT disposition phases -- a collapsed one-compartment model would
  # not produce them.
  alpha > 3 * beta,
  # The lag is honoured.
  all(ref$Cc[ref$time < p$tlag] == 0)
)
```

## The paper's own covariate-extremes calculation

This is the single most informative check in the paper. Kubota 2018 states that
if age (18-90 years), CLcr (5.8-311 mL/min), race and gender independently took
their most extreme values, "the population predicted CL/F would range from 5.17
to 12.1 L/h, which would decrease the population predicted AUCss by 25% or would
increase by 76%".

Reproducing those four numbers is an end-to-end check on the continuous
covariate terms: both centring constants (52 years, 108 mL/min), both exponents,
and the magnitude of the two categorical factors all have to be right for the
bounds to land. Note what it does **not** test: because the paper lets each
covariate "independently take extreme values", the optimum is free to pick
either level of a binary covariate, so the extremes are the same whichever way
the race and gender indicators are coded. The polarity question is settled by
the reference-subject anchor above, not here (see *Assumptions and deviations*).

```{r covariate-extremes}
cl_typ <- function(age, crcl, race_white, sexf) {
  ini <- mod$iniDf$est[match(
    c("lcl", "e_age_cl", "e_crcl_cl", "e_nonwhite_cl", "e_female_cl"),
    mod$iniDf$name
  )]
  exp(ini[1]) * (age / 52)^ini[2] * (crcl / 108)^ini[3] *
    ini[4]^(1 - race_white) * ini[5]^sexf
}

# Highest CL/F: youngest, best renal function, White, male.
cl_hi <- cl_typ(age = 18, crcl = 311.8, race_white = 1, sexf = 0)
# Lowest CL/F: oldest, worst renal function, non-White, female.
cl_lo <- cl_typ(age = 90, crcl = 5.8, race_white = 0, sexf = 1)

extremes <- tibble::tribble(
  ~Quantity,                  ~Published, ~Reproduced,
  "Highest CL/F (L/h)",          12.1,    cl_hi,
  "Lowest CL/F (L/h)",            5.17,   cl_lo,
  "AUCss change, highest CL/F (%)", -25,  100 * (9.10 / cl_hi - 1),
  "AUCss change, lowest CL/F (%)",   76,  100 * (9.10 / cl_lo - 1)
) |>
  dplyr::mutate(`Absolute difference` = abs(Reproduced - Published))

extremes |>
  dplyr::mutate(Reproduced = round(Reproduced, 3), `Absolute difference` = round(`Absolute difference`, 3)) |>
  knitr::kable(caption = "Kubota 2018 Results, covariate-extremes paragraph, reproduced from the packaged parameters.")

# Deterministic arithmetic on printed values. The published CL/F bounds carry 3
# significant figures and the AUCss percentages are printed as integers, so the
# tolerances below are the paper's own rounding, not slack.
stopifnot(
  abs(cl_hi - 12.1) < 0.01,
  abs(cl_lo - 5.17) < 0.01,
  abs(100 * (9.10 / cl_hi - 1) - (-25)) < 0.5,
  abs(100 * (9.10 / cl_lo - 1) - 76) < 0.5
)
```

## Steady-state exposure identity

Naldemedine PK is linear, so at steady state on a once-daily regimen
`AUCtau = Dose / (CL/F)` exactly. Kubota 2018 reports 21.98 ng*h/mL for the
reference subject at 0.2 mg once daily, and derived every individual AUCss used
in the exposure-response analysis the same way.

```{r ss-identity}
ev_ss <- rxode2::et(amt = 0.2, cmt = "depot", ii = 24, addl = 14) |>
  rxode2::et(seq(0, 24 * 15, by = 0.05), cmt = "central")

ss <- solve_typical(mod, ref_cov, ev_ss)
ss <- ss[!is.na(ss$Cc), ]
last <- ss[ss$time >= 24 * 14, ]
auc_tau <- sum(diff(last$time) * (utils::head(last$Cc, -1) + utils::tail(last$Cc, -1)) / 2)
auc_id <- 1000 * 0.2 / ss$cl[1]

cat(sprintf(
  "AUCtau (trapezoid over the 15th interval) = %.4f ng*h/mL\nDose / (CL/F)                           = %.4f ng*h/mL\nKubota 2018 Results                     = 21.98 ng*h/mL\n",
  auc_tau, auc_id
))

stopifnot(
  # Mass-balance identity: numerical, so tight.
  abs(auc_tau / auc_id - 1) < 0.002,
  # Against the published typical value (printed to 4 significant figures).
  abs(auc_tau - 21.98) < 0.05
)
```

# Virtual cohort

The cohort replicates the design of the phase 2b dose-finding study 1107V9221 --
patients with chronic non-cancer pain and OIC, dosed once daily under fasted
conditions at 0.1, 0.2 and 0.4 mg -- because that is the arm structure
Supplemental Table S4 reports observed steady-state exposures for. Covariates
are drawn from the pooled Table II marginal distributions and truncated to the
observed ranges.

```{r cohort}
set.seed(20181002)
rxode2::rxSetSeed(20181002)

n_per_arm <- 120L
doses <- c(0.1, 0.2, 0.4)

rtrunc <- function(n, mean, sd, lo, hi) pmin(pmax(stats::rnorm(n, mean, sd), lo), hi)

cohort <- tidyr::expand_grid(
  dose = doses,
  k = seq_len(n_per_arm)
) |>
  dplyr::mutate(
    id = dplyr::row_number(),
    arm = paste0(format(dose, trim = TRUE), " mg"),
    # Table II continuous covariates: mean, SD and observed range.
    AGE = rtrunc(dplyr::n(), 51, 14, 18, 90),
    WT = rtrunc(dplyr::n(), 79.5, 23.4, 34.4, 188.1),
    CRCL = rtrunc(dplyr::n(), 109.5, 42.4, 5.8, 311.8),
    # Table II categorical covariates.
    SEXF = stats::rbinom(dplyr::n(), 1L, 0.488),
    RACE_WHITE = stats::rbinom(dplyr::n(), 1L, 0.588),
    # Study 1107V9221 enrolled patients with chronic non-cancer pain and OIC,
    # dosed fasted: neither the healthy nor the cancer indicator is set.
    DIS_HEALTHY = 0,
    DIS_CANCER = 0,
    FED = 0
  ) |>
  dplyr::select(-k)

stopifnot(nrow(cohort) == n_per_arm * length(doses), max(table(cohort$arm)) <= 200L)

cohort |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    n = dplyr::n(),
    `Age (median)` = round(stats::median(AGE), 1),
    `Weight (median)` = round(stats::median(WT), 1),
    `CLcr (median)` = round(stats::median(CRCL), 1),
    `Female %` = round(100 * mean(SEXF), 1),
    `White %` = round(100 * mean(RACE_WHITE), 1),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Virtual cohort, drawn from the Kubota 2018 Table II marginal covariate distributions.")
```

# Simulation

Fifteen once-daily doses to reach steady state, then a dense observation grid
over the final dosing interval.

```{r simulate}
dose_rows <- cohort |>
  dplyr::transmute(
    id, arm, AGE, WT, CRCL, SEXF, RACE_WHITE, DIS_HEALTHY, DIS_CANCER, FED,
    time = 0, amt = dose, evid = 1L, cmt = "depot", ii = 24, addl = 14L
  )

obs_rows <- cohort |>
  dplyr::select(id, arm, AGE, WT, CRCL, SEXF, RACE_WHITE, DIS_HEALTHY, DIS_CANCER, FED) |>
  tidyr::crossing(time = 24 * 14 + seq(0, 24, by = 0.25)) |>
  dplyr::mutate(
    amt = NA_real_, evid = 0L,
    # The ODE STATE name, never the observable name "Cc" -- referencing an
    # algebraic observable as a compartment renumbers the compartment slots.
    cmt = "central",
    ii = 0, addl = 0L
  )

events <- dplyr::bind_rows(dose_rows, obs_rows) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

sim <- rxode2::rxSolve(mod, events, returnType = "data.frame")
if (is.null(sim$id)) sim$id <- 1L

# rxSolve drops character columns such as `arm`, so join the cohort labels back
# by id. The join must be total -- a partial match would silently shrink the
# cohort and every downstream summary with it.
sim <- sim |>
  dplyr::mutate(id = as.integer(as.character(id))) |>
  dplyr::left_join(dplyr::select(cohort, id, arm, dose), by = "id") |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(tad = time - 24 * 14)

stopifnot(
  nrow(sim) > 0L, all(sim$Cc >= 0), !anyNA(sim$Cc),
  !anyNA(sim$arm),
  dplyr::n_distinct(sim$id) == nrow(cohort)
)
```

## Steady-state concentration-time profiles

```{r fig-profiles, fig.cap = "Simulated steady-state naldemedine profiles over the final 24 h dosing interval at 0.1, 0.2 and 0.4 mg once daily (median and 5th-95th percentile band). Compare with the observed spread in Kubota 2018 Figure 1b and the prediction-corrected VPC in Figure 2."}
sim |>
  dplyr::group_by(arm, tad) |>
  dplyr::summarise(
    lo = stats::quantile(Cc, 0.05),
    md = stats::median(Cc),
    hi = stats::quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot2::ggplot(ggplot2::aes(tad, md, colour = arm, fill = arm)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::scale_y_log10() +
  ggplot2::labs(
    x = "Time after dose (h)", y = "Naldemedine concentration (ng/mL)",
    colour = "Dose", fill = "Dose"
  ) +
  ggplot2::theme_bw()
```

The profiles are dose-proportional, as the linear model requires: the bands at
0.1, 0.2 and 0.4 mg are vertical translations of one another on the log scale.

# PKNCA validation

Non-compartmental analysis over the steady-state interval, grouped by dose arm
so each arm can be compared against its own Supplemental Table S4 row.

```{r pknca}
conc_data <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, arm, time, Cc)

# Defensive time-zero anchor: PKNCA warns on every subject if the interval
# starts before the first measurement. The grid above already contains
# tad = 0, so this is a no-op here, kept so the block stays correct if the
# grid is ever coarsened.
conc_data <- conc_data |>
  dplyr::bind_rows(
    sim |>
      dplyr::filter(tad == 0) |>
      dplyr::select(id, arm, time, Cc)
  ) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

dose_data <- cohort |>
  dplyr::transmute(id, arm, time = 24 * 14, dose = dose)

o_conc <- PKNCA::PKNCAconc(conc_data, Cc ~ time | id / arm)
# PKNCAdose does not accept slash (nested) grouping, only PKNCAconc does.
o_dose <- PKNCA::PKNCAdose(dose_data, dose ~ time | id + arm)

intervals <- data.frame(
  start = 24 * 14, end = 24 * 15,
  auclast = TRUE, cmax = TRUE, cmin = TRUE, tmax = TRUE
)

o_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals))

nca_res <- as.data.frame(o_nca) |>
  dplyr::filter(PPTESTCD %in% c("auclast", "cmax", "cmin", "tmax"))

stopifnot(nrow(nca_res) > 0L, !anyNA(nca_res$PPORRES))
```

```{r pknca-summary}
nca_summary <- nca_res |>
  dplyr::group_by(arm, PPTESTCD) |>
  dplyr::summarise(
    gm = exp(mean(log(pmax(PPORRES, 1e-12)))),
    md = stats::median(PPORRES),
    cv = 100 * sqrt(exp(stats::var(log(pmax(PPORRES, 1e-12)))) - 1),
    .groups = "drop"
  )

nca_summary |>
  dplyr::mutate(
    Parameter = dplyr::recode(
      PPTESTCD,
      auclast = "AUCss (ng*h/mL)",
      cmax = "Cmax,ss (ng/mL)",
      cmin = "Cmin,ss (ng/mL)",
      tmax = "Tmax (h)"
    ),
    `Geometric mean` = round(gm, 3),
    Median = round(md, 3),
    `CV%` = round(cv, 1)
  ) |>
  dplyr::select(Arm = arm, Parameter, `Geometric mean`, Median, `CV%`) |>
  knitr::kable(caption = "Simulated steady-state exposure by dose arm.")
```

## The NCA AUC recovers Dose / (CL/F) per subject

The strongest available internal check: because the model is linear, each
subject's steady-state AUCtau must equal that subject's `Dose / (CL/F)`. This
is a per-subject identity, so it tests the covariate model, the dosing path and
the solver all at once, and it cannot be satisfied by a mis-specified
compartment structure that happens to have the right clearance.

```{r auc-identity}
subj_cl <- sim |>
  dplyr::group_by(id, arm) |>
  dplyr::summarise(cl = dplyr::first(cl), .groups = "drop")

auc_check <- nca_res |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(id, arm, auc = PPORRES) |>
  dplyr::inner_join(subj_cl, by = c("id", "arm")) |>
  dplyr::inner_join(dplyr::select(cohort, id, dose), by = "id") |>
  dplyr::mutate(
    expected = 1000 * dose / cl,
    pct_diff = 100 * (auc - expected) / expected
  )

stopifnot(nrow(auc_check) == nrow(cohort))

cat(sprintf(
  "Per-subject AUCtau vs Dose/(CL/F): median %+.3f%%, 90th percentile of |difference| %.3f%%, max %.3f%%\n",
  stats::median(auc_check$pct_diff),
  stats::quantile(abs(auc_check$pct_diff), 0.9),
  max(abs(auc_check$pct_diff))
))

stopifnot(
  # Trapezoidal error on a 0.25 h grid against an exact identity. This is
  # numerical, not a cohort statistic, but the peak of a fast-absorbing subject
  # is under-resolved by the grid, so the bound is on the centre and a robust
  # quantile rather than on the worst subject.
  abs(stats::median(auc_check$pct_diff)) < 0.5,
  stats::quantile(abs(auc_check$pct_diff), 0.9) < 2
)
```

# Comparison against published exposures

Kubota 2018 reports no non-compartmental analysis of its own. What it does
publish is Supplemental Table S4, the summary of **empirical Bayes estimated**
AUCss by study and dose arm, which is the quantity the exposure-response models
consume.

```{r nca-comparison}
reference <- tibble::tribble(
  ~arm,     ~PPTESTCD,  ~PPORRES,
  "0.1 mg", "auclast",     10.50,
  "0.2 mg", "auclast",     22.11,
  "0.4 mg", "auclast",     43.76
)

nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = reference,
  by = "arm",
  params = c("AUCss (ng*h/mL)" = "auclast"),
  label_first_column = "NCA parameter"
) |>
  knitr::kable(caption = "Simulated versus Kubota 2018 Supplemental Table S4 mean steady-state AUC for the phase 2b study 1107V9221. Values marked with a star differ by more than 20%.")
```

Two caveats govern how tightly this comparison can be gated.

* Supplemental Table S4's arms contain **9 to 10 subjects each**, and their
  AUCss values are empirical Bayes estimates shrunk toward the typical value.
  The virtual cohort here has 120 subjects per arm drawn from the *pooled*
  Table II covariate distribution, which is not the same covariate mix as those
  particular 9 patients. The comparison is therefore between a population
  prediction and a small-sample posterior summary.
* The published values are arithmetic means of a right-skewed distribution
  (Table S4 also reports geometric means of 10.2, 20.29 and 41.64), so the
  metric matters.

The deterministic anchor that *is* directly comparable -- the typical-subject
AUCss of 21.98 ng*h/mL at 0.2 mg -- was gated tightly in the *Steady-state
exposure identity* section above, and sits 0.6% below the Table S4 phase 2b
0.2 mg mean of 22.11.

```{r nca-assertions}
check <- nca_summary |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::inner_join(reference, by = c("arm", "PPTESTCD")) |>
  dplyr::mutate(pct_diff = 100 * (gm - PPORRES) / PPORRES)

check |>
  dplyr::select(arm, simulated_gm = gm, published_mean = PPORRES, pct_diff) |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable(caption = "Percent difference between the simulated geometric mean and the Supplemental Table S4 arithmetic mean AUCss.")

stopifnot(
  # Cohort-derived: gated on the centre, with headroom for the 9-10 subject
  # published arms and for the arithmetic-versus-geometric mean mismatch. A
  # mis-transcribed clearance or dose unit moves these by a factor, not by
  # tens of percent, so the gate can still go red.
  abs(stats::median(check$pct_diff)) < 20,
  max(abs(check$pct_diff)) < 25,
  # Dose proportionality is a structural property of a linear model and is
  # independent of the cohort draw.
  abs(check$gm[check$arm == "0.4 mg"] / check$gm[check$arm == "0.1 mg"] - 4) < 0.2
)
```

# Exposure-response models

```{r er-inventory}
er_names <- c(
  "Kubota_2018_naldemedine_sbm_phase2b",
  "Kubota_2018_naldemedine_sbm_phase3",
  "Kubota_2018_naldemedine_gi_mild_phase2b",
  "Kubota_2018_naldemedine_gi_moderate_phase2b",
  "Kubota_2018_naldemedine_gi_mild_phase3",
  "Kubota_2018_naldemedine_gi_moderate_phase3",
  "Kubota_2018_naldemedine_gi_severe_phase3"
)

er <- lapply(er_names, \(n) rxode2::rxode(readModelDb(n)))
names(er) <- er_names

er_coef <- function(ui, nm) ui$iniDf$est[match(nm, ui$iniDf$name)]

# Resolve each model's probability output by solving it once and reading the
# returned column names, rather than by parsing the model body.
predict_er <- function(ui, auc) {
  ev <- as.data.frame(rxode2::et(seq_along(auc) - 1))
  ev$AUC_NALD <- auc
  out <- rxode2::rxSolve(ui, ev, returnType = "data.frame")
  out[[grep("^prob_", names(out), value = TRUE)[1]]]
}

er_output <- function(ui) {
  ev <- as.data.frame(rxode2::et(0))
  ev$AUC_NALD <- 0
  out <- rxode2::rxSolve(ui, ev, returnType = "data.frame")
  grep("^prob_", names(out), value = TRUE)[1]
}

er_tab <- tibble::tibble(
  Model = sub("Kubota_2018_naldemedine_", "", er_names),
  Output = vapply(er, er_output, character(1)),
  a = vapply(er, er_coef, numeric(1), nm = "logit_ref"),
  b = vapply(er, er_coef, numeric(1), nm = "e_auc_logit")
)

er_tab |> knitr::kable(caption = "The seven landmark exposure-response models, with their Table V coefficients as packaged.")

stopifnot(length(er) == 7L, !anyNA(er_tab$a), !anyNA(er_tab$b))
```

## Source trace: every coefficient against Table V

```{r er-source-trace}
tv <- tibble::tribble(
  ~Model,                  ~Parameter, ~Published, ~`95% CI`,
  "sbm_phase2b",           "a",        -0.314,     "-0.797 to 0.159",
  "sbm_phase2b",           "b",         0.0191,    "-0.00801 to 0.0497",
  "sbm_phase3",            "a",        -0.537,     "-0.700 to -0.375",
  "sbm_phase3",            "b",         0.0194,    "0.0114 to 0.0274",
  "gi_mild_phase2b",       "a",        -1.76,      "-2.46 to -1.16",
  "gi_mild_phase2b",       "b",         0.0310,    "0.00138 to 0.0617",
  "gi_moderate_phase2b",   "a",        -2.75,      "-3.84 to -1.90",
  "gi_moderate_phase2b",   "b",         0.0254,    "-0.0170 to 0.0615",
  "gi_mild_phase3",        "a",        -1.72,      "-1.95 to -1.51",
  "gi_mild_phase3",        "b",         0.0106,    "0.000712 to 0.0202",
  "gi_moderate_phase3",    "a",        -2.75,      "-3.09 to -2.44",
  "gi_moderate_phase3",    "b",         0.0125,    "-0.00169 to 0.0259",
  "gi_severe_phase3",      "a",        -4.39,      "-5.15 to -3.77",
  "gi_severe_phase3",      "b",         0.015,     "-0.0152 to 0.0409"
)

packaged_er <- er_tab |>
  tidyr::pivot_longer(c(a, b), names_to = "Parameter", values_to = "Packaged")

er_trace <- tv |>
  dplyr::inner_join(packaged_er, by = c("Model", "Parameter")) |>
  dplyr::mutate(Difference = Packaged - Published)

er_trace |>
  dplyr::select(Model, Parameter, Published, Packaged, `95% CI`, Difference) |>
  knitr::kable(caption = "Every exposure-response coefficient against Kubota 2018 Table V.")

stopifnot(nrow(er_trace) == 14L, max(abs(er_trace$Difference)) == 0)
```

## Reproducing Table VI

Kubota 2018 Table VI tabulates the model-predicted probability at each dose for
every endpoint. Reproducing all 28 cells is an end-to-end check of the logistic
form, the coefficients and the exposure convention at once.

Note the exposure convention: the Table VI footnotes state that the 0.1 mg AUC
was "assumed to be half" and the 0.4 mg AUC "double" the 0.2 mg value, so the
0.1 and 0.4 mg columns use 11.06 / 44.22 (phase 2b) and 13.75 / 55.00
(phase 3), **not** the observed Table S4 means of 10.50 and 43.76.

```{r er-table6}
auc_p2b <- c(0, 11.06, 22.11, 44.22)
auc_p3 <- c(0, 13.75, 27.50, 55.00)

published_t6 <- tibble::tribble(
  ~Model,                ~`0 mg`, ~`0.1 mg`, ~`0.2 mg`, ~`0.4 mg`,
  "sbm_phase2b",          0.422,   0.474,     0.527,     0.630,
  "sbm_phase3",           0.369,   0.433,     0.499,     0.630,
  "gi_mild_phase2b",      0.146,   0.195,     0.254,     0.403,
  "gi_moderate_phase2b",  0.060,   0.078,     0.101,     0.165,
  "gi_mild_phase3",       0.151,   0.172,     0.193,     0.242,
  "gi_moderate_phase3",   0.060,   0.071,     0.083,     0.113,
  "gi_severe_phase3",     0.012,   0.015,     0.018,     0.027
)

simulated_t6 <- lapply(er_names, function(n) {
  short <- sub("Kubota_2018_naldemedine_", "", n)
  auc <- if (grepl("phase2b$", n)) auc_p2b else auc_p3
  stats::setNames(
    as.list(round(predict_er(er[[n]], auc), 3)),
    c("0 mg", "0.1 mg", "0.2 mg", "0.4 mg")
  ) |>
    tibble::as_tibble() |>
    dplyr::mutate(Model = short, .before = 1)
}) |>
  dplyr::bind_rows()

cmp_t6 <- published_t6 |>
  tidyr::pivot_longer(-Model, names_to = "Dose", values_to = "Published") |>
  dplyr::inner_join(
    simulated_t6 |> tidyr::pivot_longer(-Model, names_to = "Dose", values_to = "Reproduced"),
    by = c("Model", "Dose")
  ) |>
  dplyr::mutate(Difference = Reproduced - Published)

cmp_t6 |>
  tidyr::pivot_wider(id_cols = Model, names_from = Dose, values_from = Reproduced) |>
  knitr::kable(caption = "Reproduced Kubota 2018 Table VI probabilities. Compare cell by cell with the published table above.")

cat(sprintf(
  "All %d Table VI cells reproduced; largest absolute difference %.4f\n",
  nrow(cmp_t6), max(abs(cmp_t6$Difference))
))

stopifnot(
  # Deterministic algebra on published coefficients. The only tolerance needed
  # is the paper's own rounding: the intercepts are printed to 3 significant
  # figures, so a reproduced probability can land one unit off in the third
  # decimal place. The comparison is made on values already rounded to 3
  # decimals, so the bound is stated as 1.5 units in that place to keep binary
  # floating-point representation error (0.629 - 0.630 is fractionally more
  # than 0.001) from deciding the gate.
  nrow(cmp_t6) == 28L,
  max(abs(cmp_t6$Difference)) < 0.0015,
  # Measured: 20 of the 28 cells reproduce exactly and 8 sit one unit off in
  # the third decimal, which is the rounding interval of the published
  # intercepts. The bound is 10 rather than 8 so that a change of one unit in
  # the last place of one more cell does not fail the build; a genuine
  # transcription error would move cells by tens of units, not one.
  sum(abs(cmp_t6$Difference) > 1e-9) <= 10L
)
```

## Against the observed frequencies (Supplemental Table S7)

Supplemental Table S7 sets the model predictions beside the observed event
frequencies. The predictions are the Table VI values already reproduced above;
what this section adds is the observed column, which shows how well the logistic
fits describe the data they were fitted to.

```{r er-observed}
obs_pred <- tibble::tribble(
  ~Model,                ~Dose,    ~Observed, ~Predicted,
  "sbm_phase2b",         "0 mg",    39.3,      42.2,
  "sbm_phase2b",         "0.1 mg",  55.6,      47.4,
  "sbm_phase2b",         "0.2 mg",  66.7,      52.7,
  "sbm_phase2b",         "0.4 mg",  60.0,      63.0,
  "sbm_phase3",          "0 mg",    34.1,      36.9,
  "sbm_phase3",          "0.2 mg",  53.3,      49.9,
  "gi_mild_phase2b",     "0 mg",    13.1,      14.6,
  "gi_mild_phase2b",     "0.1 mg",  22.2,      19.5,
  "gi_mild_phase2b",     "0.2 mg",  44.4,      25.4,
  "gi_mild_phase2b",     "0.4 mg",  30.0,      40.3,
  "gi_moderate_phase2b", "0 mg",     3.3,       6.0,
  "gi_moderate_phase2b", "0.1 mg",  11.1,       7.8,
  "gi_moderate_phase2b", "0.2 mg",  33.3,      10.1,
  "gi_moderate_phase2b", "0.4 mg",  10.0,      16.5,
  "gi_mild_phase3",      "0 mg",    13.7,      15.1,
  "gi_mild_phase3",      "0.2 mg",  21.1,      19.3,
  "gi_moderate_phase3",  "0 mg",     5.1,       6.0,
  "gi_moderate_phase3",  "0.2 mg",   9.4,       8.3,
  "gi_severe_phase3",    "0 mg",     1.3,       1.2,
  "gi_severe_phase3",    "0.2 mg",   1.8,       1.8
)

obs_pred |>
  dplyr::mutate(`Observed - predicted` = round(Observed - Predicted, 1)) |>
  knitr::kable(caption = "Kubota 2018 Supplemental Table S7: observed frequency (%) versus model-predicted probability (%).")

# The predicted column of Table S7 must agree with the Table VI reproduction
# above -- they are the same quantity printed in two places, so this is a
# consistency check on the transcription, not a new model result.
s7_check <- obs_pred |>
  dplyr::inner_join(
    cmp_t6 |> dplyr::mutate(Reproduced100 = 100 * Reproduced),
    by = c("Model", "Dose")
  ) |>
  dplyr::mutate(d = abs(Reproduced100 - Predicted))

stopifnot(nrow(s7_check) == 20L, max(s7_check$d) <= 0.11)
```

The phase 3 fits track their observed frequencies closely. The phase 2b safety
fits do not -- at 0.2 mg the observed moderate-or-worse frequency is 33.3%
against a predicted 10.1% -- but those arms contain 9 to 10 patients each, so a
single event moves the observed proportion by 10 percentage points. Kubota 2018
says as much, and recommends the phase 3 models for prediction.

## Replicating Supplemental Figure S6

```{r er-figure-s6, fig.cap = "Replicates Kubota 2018 Supplemental Figure S6: probability of an SBM response and of a gastrointestinal disorder at each severity threshold, as a function of steady-state AUC. Vertical lines mark the mean AUCss at each dose (Supplemental Table S4 / Table VI convention)."}
auc_grid <- seq(0, 60, by = 0.5)

curves <- lapply(er_names, function(n) {
  tibble::tibble(
    Model = sub("Kubota_2018_naldemedine_", "", n),
    Phase = if (grepl("phase2b$", n)) "Phase 2b (1107V9221)" else "Phase 3 (1314V9231 + 1315V9232)",
    Endpoint = dplyr::case_when(
      grepl("^sbm", n) ~ "SBM responder",
      grepl("gi_mild", n) ~ "GI: mild or worse",
      grepl("gi_moderate", n) ~ "GI: moderate or worse",
      TRUE ~ "GI: severe"
    ),
    AUCss = auc_grid,
    Probability = predict_er(er[[n]], auc_grid)
  )
}) |>
  dplyr::bind_rows()

dose_marks <- dplyr::bind_rows(
  tibble::tibble(Phase = "Phase 2b (1107V9221)", AUCss = auc_p2b[-1]),
  tibble::tibble(Phase = "Phase 3 (1314V9231 + 1315V9232)", AUCss = auc_p3[-1])
)

ggplot2::ggplot(curves, ggplot2::aes(AUCss, Probability, colour = Endpoint)) +
  ggplot2::geom_vline(data = dose_marks, ggplot2::aes(xintercept = AUCss),
                      colour = "steelblue", linetype = "dotted") +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::facet_wrap(~Phase) +
  ggplot2::coord_cartesian(ylim = c(0, 1)) +
  ggplot2::labs(x = "Steady-state AUC (ng*h/mL)", y = "Predicted probability") +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

```{r er-shape-checks}
# Structural properties of the fitted set, all deterministic.
shape <- curves |>
  dplyr::group_by(Model) |>
  dplyr::summarise(
    increasing = Probability[dplyr::n()] > Probability[1],
    in_unit = all(Probability >= 0 & Probability <= 1),
    .groups = "drop"
  )

stopifnot(all(shape$increasing), all(shape$in_unit))

# The three severity thresholds are CUMULATIVE, so within a study set the
# mild-or-worse curve must dominate moderate-or-worse, which must dominate
# severe. The authors fitted them independently, with no proportional-odds
# constraint, so this ordering is a property of the estimates rather than of
# the parameterisation -- worth checking rather than assuming.
ord <- curves |>
  dplyr::filter(grepl("^gi_", Model), grepl("Phase 3", Phase)) |>
  dplyr::select(Model, AUCss, Probability) |>
  tidyr::pivot_wider(names_from = Model, values_from = Probability)

stopifnot(
  all(ord$gi_mild_phase3 >= ord$gi_moderate_phase3),
  all(ord$gi_moderate_phase3 >= ord$gi_severe_phase3)
)
```

## The paper's headline safety claim

Kubota 2018 concludes that at 0.2 mg "the probability of the occurrence of
severe gastrointestinal disorders was predicted to be less than 3%", and that
the same holds at 0.1 and 0.4 mg.

```{r er-headline}
severe <- predict_er(er[["Kubota_2018_naldemedine_gi_severe_phase3"]], auc_p3)
names(severe) <- c("0 mg", "0.1 mg", "0.2 mg", "0.4 mg")
print(round(100 * severe, 2))

sbm_p3 <- predict_er(er[["Kubota_2018_naldemedine_sbm_phase3"]], auc_p3[3])
cat(sprintf("Predicted SBM responder probability at 0.2 mg (phase 3): %.1f%%\n", 100 * sbm_p3))

stopifnot(
  # "less than 3%" at every studied dose -- deterministic, from the paper's
  # own predicted exposures.
  all(severe < 0.03),
  # "approximately 50%" SBM responders at 0.2 mg.
  abs(sbm_p3 - 0.50) < 0.02
)
```

## End to end: the PK model feeds the exposure-response models

The two layers are connected in the source analysis by the individual AUCss.
Here the simulated cohort's NCA exposures are pushed through the phase 3
efficacy and severe-safety models, which is the operation Kubota 2018 performed
with empirical Bayes estimates.

```{r er-endtoend}
endtoend <- auc_check |>
  dplyr::select(id, arm, AUCss = auc) |>
  dplyr::mutate(
    p_sbm = predict_er(er[["Kubota_2018_naldemedine_sbm_phase3"]], AUCss),
    p_severe = predict_er(er[["Kubota_2018_naldemedine_gi_severe_phase3"]], AUCss)
  )

endtoend |>
  dplyr::group_by(Arm = arm) |>
  dplyr::summarise(
    `Median AUCss` = round(stats::median(AUCss), 2),
    `Mean P(SBM responder)` = round(mean(p_sbm), 3),
    `Mean P(severe GI)` = round(mean(p_severe), 4),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Cohort-average predicted probabilities from PK-simulated steady-state exposures, phase 3 models.")

stopifnot(
  nrow(endtoend) == nrow(cohort),
  all(endtoend$p_sbm > 0 & endtoend$p_sbm < 1),
  # The efficacy relationship is shallow by construction; the claim under test
  # is that quadrupling the dose moves the responder probability by a modest
  # amount, not that it is flat. Cohort-derived, so gated on the mean with
  # headroom rather than on any subject.
  mean(endtoend$p_sbm[endtoend$arm == "0.4 mg"]) > mean(endtoend$p_sbm[endtoend$arm == "0.1 mg"]),
  mean(endtoend$p_severe) < 0.05
)
```

This is the quantitative basis for the paper's conclusion. The covariate
extremes move AUCss by -25% to +76% (checked above), and over that exposure
range the responder probability changes by only a few percentage points while
the severe-event probability stays under 3%. That is why Kubota 2018 concludes
no covariate-driven dose adjustment is required.

# Assumptions and deviations

## The race indicator's polarity is inverted relative to its name

This is the one reading in the paper that could plausibly have been taken the
other way, and it is load-bearing: it decides which group carries a 13%
clearance difference.

Kubota 2018's Table III equation multiplies CL/F by `THETA(4)^White`, which
reads as "0.870 applies to White subjects". The Table II footnote says the
opposite: "White = 1 for non-White, White = 0 for White". The indicator is named
for the category it *excludes*.

The footnote is correct, and the paper settles it in its own Results text. The
reference population -- the subject at which every indicator is zero, and whose
CL/F is therefore THETA(1) unmodified -- is described as "52-year-old, 76 kg
male, white, non-cancer, OIC patients with CLcr of 108 ml/min, administered
under fasted condition", with CL/F = 9.10 L/h. That subject is **White** and
carries no race factor, so `White = 0` for White subjects and the 0.870 applies
to non-White. The reference-subject anchor check above is the mechanical gate on
this: it solves a `RACE_WHITE = 1` subject and requires CL/F within 0.05% of
9.10. Under the equation-as-printed reading the same subject would return
9.10 x 0.870 = 7.92 L/h, a 13% miss that blows the gate by a factor of 260.

The covariate-extremes check is *not* what decides this, despite being the more
tempting argument: the paper's extremes let each covariate vary independently,
so the free optimum simply attaches the 0.870 to whichever race label minimises
CL/F, and both codings reproduce 5.17 and 12.1 L/h. It pins the continuous
terms, not the polarity.

The model file therefore carries the published 0.870 unchanged and raises it to
the power `(1 - RACE_WHITE)`, so the canonical `RACE_WHITE` column keeps its
usual polarity and the inversion is visible at the call site.

## Health status is carried on two ratified canonical columns

The source codes a three-level health status (healthy / chronic non-cancer pain
with OIC / cancer with OIC) with two indicator variables, in which healthy
subjects are the cell where both vanish. Rather than mint a new canonical for
"chronic non-cancer pain with OIC", the model carries the ratified `DIS_HEALTHY`
and `DIS_CANCER` columns and recovers the source's `non-Cancer` indicator inside
`model()` as `(1 - DIS_HEALTHY) * (1 - DIS_CANCER)`. Encode a subject as healthy
with `DIS_HEALTHY = 1, DIS_CANCER = 0`; as chronic-non-cancer-pain OIC with both
0; as cancer OIC with `DIS_CANCER = 1`.

## The body-weight exponent on Vc/F is fixed at 1, not estimated

Table III writes the age and creatinine-clearance terms as powers but writes the
weight term as `* (Body weight/76)` with no exponent, so the exponent was held
at 1 rather than estimated. The model file encodes this as `e_wt_vc <- fixed(1)`
so the distinction between a fixed structural assumption and an estimate is not
lost.

## Creatinine clearance is raw Cockcroft-Gault, not BSA-normalised

The `CRCL` canonical's primary definition is BSA-normalised mL/min/1.73 m^2,
but Kubota 2018 uses a raw Cockcroft-Gault clearance in mL/min (Methods, citing
reference 12), centred on the cohort median of 108 mL/min. Supplying a
BSA-normalised value to this model biases CL/F by `(ratio)^0.0739`. The exponent
is small, so the error is modest -- but it is still an error.

## The creatinine-clearance effect's confidence interval spans zero

The exponent 0.0739 has a 95% CI of -0.0133 to 0.161. Kubota 2018 retained the
term despite this, because a dedicated renal-impairment study had shown a real
AUC difference. That is a source-stated modelling decision and is reproduced
unchanged rather than simplified away.

## Inter-individual variability is diagonal

Kubota 2018 reports (Supplemental Figure S3) that "there was no clear
correlation between the inter-individual variabilities", and publishes no
off-diagonal elements, so the omega matrix is diagonal. Inter-individual
variability on the absorption lag time was removed by the authors to allow the
estimation and covariance routines to converge, so the model carries no
`etaltlag`.

## The IIV on Ka is very large

The published Ka CV of 161.2% is faithfully encoded, and it makes the simulated
absorption phase extremely variable -- a meaningful minority of subjects
absorb slowly enough to be flip-flop. This is what the paper reports (with 32.6%
shrinkage), not an artefact of the encoding, but it is why the NCA checks here
gate AUC and its per-subject identity rather than Tmax or a terminal half-life,
neither of which is stable across such a cohort.

## The exposure-response models carry a placeholder residual error

The source fits are binomial logistic regressions in SAS with a Bernoulli
likelihood, which has no residual error parameter and no random effects. Each
exposure-response model file exposes a tiny fixed additive residual
(`addSd_prob_* <- fixed(0.001)`) purely so the nlmixr2 observation machinery has
an error model to attach to the typical-value probability. **That value is not
from the source** and must not be interpreted as a published quantity.

## Six of the seven exposure-response slopes have confidence intervals spanning zero

Only `sbm_phase3` and `gi_mild_phase3` have slope confidence intervals excluding
zero. Kubota 2018 is explicit about this limitation: the exposure range studied
spans only half to twice the 0.2 mg clinical dose, and the event counts at the
stricter severity thresholds are small. The paper's own conclusion is that the
models "can be used to interpret the exposure response relationship" in these
studies but that "it is inappropriate to generalize them". The models are
packaged as fitted; users extrapolating outside roughly 5 to 90 ng*h/mL should
read that caveat first.

## The exposure convention differs between Table VI and Table S4

Table VI's 0.1 and 0.4 mg columns use AUCss values *assumed* to be half and
double the 0.2 mg value, not the observed means from Supplemental Table S4
(11.06 versus 10.50, and 44.22 versus 43.76 in phase 2b). Reproducing Table VI
requires the assumed values; characterising the fitted population requires the
observed ones. Both are recorded in the `AUC_NALD` register entry.

## Equation 3's raster is illegible in every available copy

The published PDF and the publisher's supplementary figure files both render
Eq. (3) at a resolution too low to read. Its form was recovered from two
independent printed sources instead: the Table VI footnotes in the PDF text
layer, which spell it out as
`Probability = 1 / [1 + exp(-a - b x AUCss)]`, and the reproduction of all 28
Table VI cells above, which the same form satisfies to the last printed decimal.
No value was digitised from a figure anywhere in this extraction.

## Not extracted

The base model of Supplemental Table S3 is a model-development intermediate and
is not packaged, per the repository's base-versus-final policy. The covariates
screened but not retained are recorded in the model file's
`covariatesDataExcluded` list for provenance.
