---
title: "Kuchimanchi_2018_evolocumab"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Kuchimanchi_2018_evolocumab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
```

```{r setup}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Evolocumab population PK replication (Kuchimanchi 2018)

Kuchimanchi et al. (2018) characterised the population pharmacokinetics of
evolocumab, a fully human IgG2 monoclonal antibody against proprotein
convertase subtilisin/kexin type 9 (PCSK9), in 3414 subjects pooled across
11 clinical studies spanning phase 1, 2, and 3 (healthy volunteers plus
patients with hypercholesterolemia, including heterozygous familial
hypercholesterolemia and statin-intolerance cohorts). The final model is a
one-compartment model with first-order SC absorption from a depot compartment
and parallel linear plus Michaelis–Menten (target-mediated) elimination from
the central compartment. Body weight entered as a power effect on CL, V, and
V~max~; female sex multiplied V; statin monotherapy, ezetimibe use
(functionally the statin + ezetimibe combination-therapy indicator), and
baseline PCSK9 modified V~max~. The absorption rate constant (k~a~), SC
bioavailability (F), Michaelis–Menten constant (k~m~), and V~max~ typical
value were fixed in the updated phase 3 population PK model.

This vignette reproduces typical-value concentration–time profiles for the
two commercial regimens (140 mg SC Q2W and 420 mg SC QM), documents the
parameter provenance in a source-trace table, and validates the simulated
NCA (PKNCA) steady-state exposures against the published mean C~max~
values.

# Model and source

* Citation: Kuchimanchi M, Monine M, Kandadi Muralidharan K, Woodhead JL,
  Horner TJ. *Population pharmacokinetics and exposure–response modeling
  and simulation for evolocumab in healthy volunteers and patients with
  hypercholesterolemia.* J Pharmacokinet Pharmacodyn. 2019;46(2):133–148.
* Article:
  [doi:10.1007/s10928-018-9592-y](https://doi.org/10.1007/s10928-018-9592-y).
* No errata were identified (PubMed search `Kuchimanchi 2018 evolocumab
  erratum`, 2026-04-24, returned no results).

# Population

Kuchimanchi 2018 Table 2 (phase 1, 2, and 3 pooled column; N = 3414):

| Field              | Value                                                                |
|--------------------|----------------------------------------------------------------------|
| N subjects         | 3414 (receiving evolocumab; pooled from 5474 across all arms)        |
| N observations     | 16 179 evolocumab concentrations                                     |
| N studies          | 11 (1 phase 1a, 1 phase 1b, 4 phase 2, 5 phase 3)                    |
| Age                | 18–80 years (mean 57)                                                |
| Weight             | 41–175 kg (mean 84.2)                                                |
| Sex                | 50% female / 50% male                                                |
| Race / ethnicity   | 87% White, 7% Black, 4% Asian, 2% other                              |
| Disease state      | Healthy volunteers (phase 1a) + hypercholesterolemia patients        |
| Baseline PCSK9     | Mean 402 ng/mL (range 15.5–1233)                                     |
| Baseline albumin   | Mean 4.3 g/dL (range 2.6–5.6)                                        |
| HeFH (%)           | 9%                                                                   |
| Diabetes (%)       | 11%                                                                  |
| Statin (any) (%)   | ~72%                                                                 |
| Ezetimibe (%)      | 12%                                                                  |
| Dose range         | 7–420 mg IV or SC across single- and multiple-dose regimens          |
| Phase 3 regimens   | 140 mg SC Q2W and 420 mg SC QM                                       |
| Reference patient  | 84 kg male, no lipid-lowering meds, baseline PCSK9 = 425 ng/mL       |

The population metadata is also available programmatically via
`readModelDb("Kuchimanchi_2018_evolocumab")$population`.

# Source trace

Every numeric value in the model file
`inst/modeldb/specificDrugs/Kuchimanchi_2018_evolocumab.R` is sourced from
Kuchimanchi 2018. Final estimates are the **updated phase 3 population PK
model** column in Table 3; V~max~ and k~m~ were fixed in that column from
the phase 1+2 run because phase 3 used only two dose regimens.

| Quantity                                   | Source location                         | Value used                                  |
|--------------------------------------------|-----------------------------------------|---------------------------------------------|
| Structural model (1-cmt, linear + MM)      | Methods § PopPK analysis / Figure 1a    | Depot → central, CL + V·V~max~·C/(k~m~+C)   |
| F (SC bioavailability)                     | Table 3                                 | 0.72 (FIXED)                                |
| k~a~                                       | Table 3                                 | 0.319 day^-1^ (FIXED)                       |
| CL                                         | Table 3                                 | 0.105 L/day                                 |
| V                                          | Table 3                                 | 5.18 L                                      |
| V~max~                                     | Table 3                                 | 9.85 nM/day (FIXED)                         |
| k~m~                                       | Table 3                                 | 27.3 nM (FIXED)                             |
| Reference body weight                      | Methods, reference-patient definition   | 84 kg (male)                                |
| WT exponent on CL                          | Table 3                                 | 0.276                                       |
| WT exponent on V                           | Table 3                                 | 1.04                                        |
| Female exponent on V                       | Table 3                                 | 1.11                                        |
| WT exponent on V~max~                      | Table 3                                 | 0.145                                       |
| Statin (monotherapy) exponent on V~max~    | Table 3                                 | 1.13                                        |
| Statin + ezetimibe exponent on V~max~      | Table 3                                 | 1.20                                        |
| Baseline PCSK9 exponent on V~max~          | Table 3                                 | 0.194                                       |
| Reference PCSK9                            | Methods, reference-patient definition   | 425 ng/mL (= 5.9 nM)                        |
| IIV on CL                                  | Table 3                                 | 54.3% CV                                    |
| IIV on V                                   | Table 3                                 | 28.3% CV                                    |
| IIV on V~max~                              | Table 3                                 | 31.1% CV                                    |
| IIV on k~a~                                | Table 3                                 | 74.6% CV (FIXED)                            |
| IIV on k~m~                                | Table 3                                 | 0% (FIXED)                                  |
| Full-block CL/V/V~max~ omega structure     | Table 3 note; Table 4 referenced but    | Correlations not published — diagonal used  |
|                                            | actually E-R parameters                 | (see Assumptions and deviations)            |
| Proportional residual error                | Table 3                                 | 0.282 (28.2% CV)                            |
| Additive residual error                    | Table 3                                 | 5.41 nM (= 0.767 µg/mL at MW 141.8 kDa)     |
| Evolocumab molecular weight                | FDA-approved Repatha label              | 141 800 g/mol                               |

PCSK9 MW (~72 kDa), used only to cross-check the paper's nM↔ng/mL
conversion (5.9 nM ≈ 425 ng/mL), is not a model parameter.

# Virtual cohort

The original patient-level data are not public. To replicate the paper's
reference-patient C~max~ we simulate a small typical-value cohort at the
two commercial regimens (140 mg SC Q2W and 420 mg SC QM) with the model's
random effects zeroed (`rxode2::zeroRe()`), i.e. a single "average" subject
per regimen matching the Methods-section reference definition (84 kg male,
no lipid-lowering medication, baseline PCSK9 = 425 ng/mL).

```{r cohort}
set.seed(20260424)

reference_covariates <- tibble::tibble(
  WT          = 84,
  SEXF        = 0L,
  STATIN_MONO = 0L,
  EZE         = 0L,
  PCSK9       = 425
)

make_regimen <- function(amt, ii, addl, label) {
  ev <- rxode2::et(amt = amt, cmt = "depot", ii = ii, addl = addl) |>
    rxode2::et(seq(0, 12 * 14, by = 0.5))
  df <- as.data.frame(ev)
  df$id          <- 1L
  df$WT          <- reference_covariates$WT
  df$SEXF        <- reference_covariates$SEXF
  df$STATIN_MONO <- reference_covariates$STATIN_MONO
  df$EZE         <- reference_covariates$EZE
  df$PCSK9       <- reference_covariates$PCSK9
  df$regimen     <- label
  df
}

events <- dplyr::bind_rows(
  make_regimen(amt = 140, ii = 14, addl = 11, label = "140 mg SC Q2W"),
  make_regimen(amt = 420, ii = 28, addl = 5,  label = "420 mg SC QM")  |>
    dplyr::mutate(id = 2L)
)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod <- readModelDb("Kuchimanchi_2018_evolocumab")
ui  <- rxode2::rxode(mod)

# Typical-value prediction: no between-subject variability and no residual error.
mod_typical <- rxode2::zeroRe(ui)
sim <- rxode2::rxSolve(mod_typical, events = events, keep = c("regimen"))
```

# Replicate published figures

```{r figure-4a, fig.cap = "Typical serum evolocumab concentration time-course after 140 mg SC Q2W and 420 mg SC QM (reference patient). Replicates the shape and range of Figure 4a/4c of Kuchimanchi 2018."}
ggplot(sim, aes(time, Cc, colour = regimen)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time (day)", y = "Serum evolocumab (µg/mL)",
       colour = "Regimen",
       title = "Typical serum evolocumab concentration-time profiles",
       caption = "Reference patient: 84 kg male, no lipid-lowering Rx, PCSK9 = 425 ng/mL.") +
  theme_minimal(base_size = 11)
```

```{r figure-4a-log, fig.cap = "Same simulation on a log scale, showing the dose-proportional scaling between Q2W and QM trough concentrations."}
ggplot(sim, aes(time, Cc, colour = regimen)) +
  geom_line(linewidth = 0.7) +
  scale_y_log10() +
  labs(x = "Time (day)", y = "Serum evolocumab (µg/mL, log)",
       colour = "Regimen") +
  theme_minimal(base_size = 11)
```

# PKNCA validation

Use PKNCA to compute steady-state AUC, C~max~, T~max~, C~min~, and C~avg~
over the last dosing interval of each regimen (days 154–168 for Q2W,
days 140–168 for QM). We keep observations between the final dose and
the end of simulation, one per regimen, so NCA runs on the steady-state
window.

```{r pknca}
tau_by_regimen <- c("140 mg SC Q2W" = 14, "420 mg SC QM" = 28)

sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(
    tau       = tau_by_regimen[regimen],
    dose_last = 12 * 14 - tau
  ) |>
  dplyr::filter(time >= dose_last) |>
  dplyr::mutate(tad = time - dose_last) |>
  dplyr::select(id, time, tad, Cc, regimen)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::group_by(id, regimen) |>
  dplyr::slice_tail(n = 1) |>     # final dose only
  dplyr::ungroup() |>
  dplyr::select(id, time, amt, regimen)

conc_obj <- PKNCA::PKNCAconc(
  sim_nca, Cc ~ time | regimen + id,
  concu = "ug/mL", timeu = "day"
)
dose_obj <- PKNCA::PKNCAdose(
  dose_df, amt ~ time | regimen + id,
  doseu = "mg"
)

intervals <- data.frame(
  start   = c(12 * 14 - 14, 12 * 14 - 28),
  end     = c(12 * 14,      12 * 14),
  cmax    = TRUE,
  tmax    = TRUE,
  cmin    = TRUE,
  auclast = TRUE,
  cav     = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)
nca_tbl <- as.data.frame(nca_res$result) |>
  dplyr::select(regimen, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
knitr::kable(nca_tbl,
             caption = "Steady-state NCA parameters from the simulated reference patient.")
```

## Comparison against published C~max~

Kuchimanchi 2018 Results, paragraph following Table 3, reports mean
observed unbound evolocumab C~max~ of **18.6 µg/mL** after 140 mg SC Q2W
and **59 µg/mL** after 420 mg SC QM. These values are population means
observed across the phase 1–3 dataset (i.e., they include
between-subject variability and covariate effects that differ from the
Methods-defined reference patient), so an exact match is not expected
for the typical-value simulation here; agreement within ~10% would be
a reasonable sanity check. The T~max~ of the SC profile is not reported
as a typical value in the paper; the simulated T~max~ is driven by the
fixed k~a~ = 0.319 day^-1^ and the regimen-specific dose accumulation.

```{r cmax-comparison}
cmax_obs <- tibble::tibble(
  regimen = c("140 mg SC Q2W", "420 mg SC QM"),
  cmax_paper_ugmL = c(18.6, 59)
)
cmax_sim <- sim |>
  dplyr::filter(time >= 12 * 14 - 28) |>  # within last QM interval window
  dplyr::group_by(regimen) |>
  dplyr::filter(time >= (12 * 14 - tau_by_regimen[regimen[1]])) |>
  dplyr::summarise(cmax_sim_ugmL = max(Cc), .groups = "drop")
cmax_cmp <- dplyr::left_join(cmax_obs, cmax_sim, by = "regimen") |>
  dplyr::mutate(pct_diff = round(100 * (cmax_sim_ugmL - cmax_paper_ugmL)
                                  / cmax_paper_ugmL, 1))
knitr::kable(cmax_cmp,
  caption = "Simulated (typical-value) steady-state C~max~ vs published mean observed C~max~.")
```

# Assumptions and deviations

- **IIV block matrix approximated as diagonal.** Kuchimanchi 2018 states
  that CL, V, and V~max~ share a full-block variance matrix and refers
  the reader to "Table 4" for the inter-parameter correlations. However,
  the paper's Table 4 is the exposure–response model parameter table and
  does not list the population-PK omega-block correlations; the
  correlations are not published anywhere in the article or supplement.
  The model file therefore encodes independent diagonal IIV on CL, V,
  and V~max~. This understates the individual-level covariance structure
  relative to the published fit (in particular the high CL–V~max~
  correlation that the paper calls out as influencing body-weight
  covariate-effect precision) but has no effect on typical-value
  (`rxode2::zeroRe()`) simulation or on population-level summaries.
- **Exposure–response (E~max~) model not packaged.** The paper additionally
  reports an E~max~-type exposure–response model (Table 4) linking
  AUC~wk8–12~ to LDL-C at the mean of weeks 10 and 12. That model is a
  cross-sectional algebraic relationship rather than a dynamic PK/PD
  ODE, so it is not a natural fit for the nlmixr2lib library. Only the
  population PK model is packaged here.
- **Molecular weight** used to convert between the paper's target-unit
  scale (nM) and the model's concentration scale (µg/mL) is 141 800 g/mol
  (evolocumab, FDA-approved Repatha label). The paper does not state the
  molecular weight explicitly; the same value reproduces the paper's
  5.9 nM ≈ 425 ng/mL reference-PCSK9 relationship within rounding when
  the PCSK9 molecular weight (~72 kDa) is applied to PCSK9.
- **Reference patient characteristics** used in the virtual cohort are
  the Methods-section reference definition: 84 kg male, not on
  lipid-lowering medication, baseline PCSK9 = 425 ng/mL. Body weight
  distribution and PCSK9 distribution in the broader trial population
  (which shifts observed mean C~max~ slightly relative to the reference
  patient) are not reproduced in this typical-value simulation.
- **Time-fixed covariates.** All covariates in the model file
  (`WT`, `SEXF`, `STATIN_MONO`, `EZE`, `PCSK9`) are treated as baseline
  time-fixed. Kuchimanchi 2018's dataset used baseline values for these
  covariates; concomitant-medication status was required to be stable
  (>4 weeks of administration before study day 1) in the
  exposure–response analysis, though the popPK analysis used any
  duration of administration.
