---
title: "Quinidine in Thoroughbred horses (Kuroda 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quinidine in Thoroughbred horses (Kuroda 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Kuroda_2024_quinidine_horse"))$reference`
* Article: <https://doi.org/10.3389/fvets.2024.1454342>
* Open-access full text: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC11493839/>

Kuroda 2024 is a population pharmacokinetic analysis of quinidine (QND) in
Thoroughbred racehorses, undertaken to replace the long-standing empirical
dosing regimen for atrial fibrillation (AF) with one chosen from the
distribution of plasma concentrations rather than from its mean. The model is a
three-compartment disposition system with first-order oral absorption and a
bioavailability factor, fitted in Phoenix WinNonlin/NLME 8.4 with the QRPEM
engine.

Two features of the paper drive everything below.

First, **every structural parameter is reported per kilogram of body weight**
(L/kg, L/kg/h in Table 1). The packaged model therefore multiplies each volume
and clearance by `WT` with a fixed exponent of 1, which is precisely what the
per-kilogram normalisation means. A useful consequence: because the doses are
also prescribed per kilogram, body weight cancels exactly out of the predicted
concentration. That is the mechanical reason the paper's covariate search found
no body-weight effect, and it is asserted explicitly below.

Second, **all doses and parameters are expressed as QND base.** The paper
administers two different salts and converts both with stated factors: 1.206 for
quinidine sulfate dihydrate (the oral form) and 1.168 for quinidine
hydrochloride monohydrate (the intravenous form). Every dose in this vignette is
converted to base before it reaches the model, and the salt-equivalent figure is
carried alongside so the tables can be read against the paper's.

```{r model}
mod <- readModelDb("Kuroda_2024_quinidine_horse")
ui <- rxode2::rxode(mod)

# Salt-to-base conversion factors (Kuroda 2024 section 2.5).
SALT_SULFATE_DIHYDRATE <- 1.206
SALT_HCL_MONOHYDRATE <- 1.168
as_base_sulfate <- function(mg_per_kg_salt) mg_per_kg_salt / SALT_SULFATE_DIHYDRATE
as_salt_sulfate <- function(mg_per_kg_base) mg_per_kg_base * SALT_SULFATE_DIHYDRATE

# Typical values, read back out of the packaged model so the checks below
# cannot drift away from the ini() block.
tv <- setNames(ui$iniDf$est, ui$iniDf$name)
TV <- list(
  vc  = exp(tv[["lvc"]]),  vp = exp(tv[["lvp"]]), vp2 = exp(tv[["lvp2"]]),
  cl  = exp(tv[["lcl"]]),  q  = exp(tv[["lq"]]),  q2  = exp(tv[["lq2"]]),
  ka  = exp(tv[["lka"]]),  fdepot = plogis(tv[["logitfdepot"]])
)
str(TV)
```

# Population

Twenty-nine Thoroughbred horses contributed plasma quinidine data (Kuroda 2024
section 2.1). Ten were healthy: six received both a single 5 mg/kg intravenous
dose of quinidine hydrochloride monohydrate (4.28 mg/kg base, as a 5-minute
infusion into the right jugular vein) and a single 20 mg/kg oral dose of
quinidine sulfate dihydrate (16.58 mg/kg base, by nasogastric tube) in a
two-way crossover with a two-week washout, and four received the same 20 mg/kg
oral dose twice, six hours apart. These ten horses were 2-7 years old, weighed
473-563 kg, and were five stallions and five mares.

The remaining nineteen were racehorses in treatment for naturally occurring
atrial fibrillation, sampled opportunistically during therapy at
veterinarian-chosen doses between 9.3 and 30.6 mg/kg quinidine sulfate dihydrate
(7.7-25.4 mg/kg base) by nasogastric tube. They were 2-10 years old, weighed
430-540 kg, and were twelve stallions and seven mares. Eighteen of the nineteen
converted to sinus rhythm.

The paper's clinical anchor comes from this second group: the median plasma
quinidine concentration at conversion to sinus rhythm was 2.0 ug/mL
(range 0.5-2.7, from the thirteen horses whose conversion fell inside the
sampling window), while the median concentration at which adverse effects
occurred was 3.8 ug/mL (range 1.6-5.1). Kuroda 2024 therefore proposes
**2.0-3.8 ug/mL** as the therapeutic window, narrower at the top than the
2-5 ug/mL previously recommended by Reef et al. (1995).

The same information is available programmatically:

```{r population}
pop <- readModelDb("Kuroda_2024_quinidine_horse")()$population
pop[c("species", "n_subjects", "age_range", "weight_range", "disease_state")]
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Kuroda_2024_quinidine_horse.R` carries an in-file
comment naming its source. They are collected here for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lvc` (V1) | 0.63 L/kg | Table 1, primary structural parameters |
| `lvp` (V2) | 0.59 L/kg | Table 1 |
| `lvp2` (V3) | 3.68 L/kg | Table 1 |
| `lcl` (CL) | 0.49 L/kg/h | Table 1 |
| `lq` (CL2) | 2.87 L/kg/h | Table 1 |
| `lq2` (CL3) | 2.44 L/kg/h | Table 1 |
| `lka` (Kabs) | 1.00 1/h | Table 1 |
| `logitfdepot` (F) | 36.4%, held on the logit scale | Table 1 + its footnote "For F, an ilogit transformation was used to prevent estimates higher than 100%" |
| `etalvc` ... `etalogitfdepot` | BSV% 40.5 / 37.8 / 28.5 / 25.6 / 74.9 / 47.3 / 94.3 / 33.1 | Table 1 BSV% column, converted with section 2.5 Equation 2, `omega^2 = log(1 + (BSV%/100)^2)` |
| `propSdIv` | 0.0594 | Table 1, CMultStdev0 (proportional, IV) |
| `propSdOral` | 0.1571 | Table 1, CMultStdev1 (proportional, PO) |
| `addSdIv` | 0.0338 ug/mL | Table 1, Stdev0 (additive, IV); units read as ug/mL, not the printed ug/L (see Errata) |
| `addSdOral` | 0.0479 ug/mL | Table 1, Stdev1 (additive, PO); same unit correction |
| Exponential BSV, `theta_i = theta_tv * exp(eta_i)` | n/a | Section 2.5 Equation 1 |
| Combined residual, `C = f(theta, t) * (1 + eps1) + eps2` | n/a | Section 2.5 Equation 4 |
| Three-compartment structure, CL/V parameterisation, Kabs and F added for PO | n/a | Section 2.5, paragraph 1 |
| Per-kilogram scaling of all volumes and clearances | n/a | Table 1 units column (L/kg, L/kg/h) |
| Salt factors 1.206 / 1.168 | n/a | Section 2.5, paragraph 1 |
| Loading dose `= C_target * Vss / F` | n/a | Section 2.5 Equations 7-8, values in Table 2 |
| Maintenance rate `= C_target * CL / F` | n/a | Section 2.5 Equations 7-8, values in Table 2 |
| Therapeutic window 2.0-3.8 ug/mL | n/a | Results paragraph 3 |

# Structural checks against the published secondary parameters

Table 1 also reports six *secondary* parameters, which are deterministic
functions of the six primary disposition parameters. They are the sharpest
available check on the transcription: a single mis-keyed volume or clearance
moves at least one of them well outside rounding. Nothing here is simulated, so
these bounds are tight by design.

```{r closed-form}
# Steady-state volume of distribution is the sum of the three volumes.
vss <- TV$vc + TV$vp + TV$vp2

# Mean residence time.
mrt <- vss / TV$cl

# Disposition half-lives are log(2) over the negated eigenvalues of the
# three-compartment rate matrix, built from the micro-constants.
k10 <- TV$cl / TV$vc
k12 <- TV$q  / TV$vc; k21 <- TV$q  / TV$vp
k13 <- TV$q2 / TV$vc; k31 <- TV$q2 / TV$vp2
rate_matrix <- matrix(
  c(-(k10 + k12 + k13), k21,  k31,
    k12,                -k21, 0,
    k13,                0,    -k31),
  nrow = 3, byrow = TRUE
)
half_lives <- sort(log(2) / -Re(eigen(rate_matrix)$values))

secondary <- tibble::tibble(
  parameter = c("Half_life_alpha", "Half_life_Beta", "Half_life_Gamma",
                "Absorption_Half_life", "Vss", "MRT"),
  units     = c("h", "h", "h", "h", "L/kg", "h"),
  model     = c(half_lives, log(2) / TV$ka, vss, mrt),
  published = c(0.06, 0.31, 7.76, 0.69, 4.90, 10.12)
) |>
  mutate(pct_diff = 100 * (model - published) / published)

secondary |>
  mutate(across(c(model, published, pct_diff), \(x) round(x, 3))) |>
  rename("Secondary parameter" = parameter, "Units" = units,
         "Model" = model, "Kuroda 2024 Table 1" = published,
         "% difference" = pct_diff) |>
  knitr::kable(caption = "Model-derived secondary parameters vs Kuroda 2024 Table 1.")

# The three half-lives and the absorption half-life are published to two
# decimals, so rounding alone permits a few percent on the 0.06 h alpha phase;
# Vss is an exact sum and is held much tighter. MRT is the one row where the
# paper's value (10.12 h) is a bootstrap median of INDIVIDUAL MRTs rather than
# the typical-value ratio Vss/CL (10.00 h), which accounts for its 1.2%.
stopifnot(
  abs(secondary$pct_diff[secondary$parameter == "Vss"]) < 0.5,
  abs(secondary$pct_diff[secondary$parameter == "MRT"]) < 2,
  abs(secondary$pct_diff[secondary$parameter == "Half_life_Gamma"]) < 1,
  max(abs(secondary$pct_diff)) < 8
)
```

The gamma (terminal) half-life reproduces to 0.03%, and `Vss` is exact. The
alpha half-life carries the largest relative difference simply because 0.056 h
rounds to 0.06.

## Table 2: the paper's own dosing equations

Section 2.5 Equations 7 and 8 compute the loading dose and the maintenance rate
needed to reach a target plasma concentration:

* loading dose `= C_target * Vss / F`
* maintenance rate `= C_target * CL / F`

Table 2 tabulates both for seven target concentrations, in QND base and in
quinidine sulfate dihydrate. Reproducing all fourteen numbers from the packaged
parameters exercises `Vss`, `CL` and `F` jointly, including the salt conversion.

```{r table-2}
targets <- c(0.5, 1.0, 1.5, 2.0, 2.5, 2.9, 3.5)

table2 <- tibble::tibble(
  target       = targets,
  ld_base      = targets * vss / TV$fdepot,
  md_base      = targets * TV$cl / TV$fdepot
) |>
  mutate(
    ld_salt = as_salt_sulfate(ld_base),
    md_salt = as_salt_sulfate(md_base),
    # Kuroda 2024 Table 2, transcribed.
    pub_ld_base = c(6.72, 13.43, 20.15, 26.86, 33.58, 38.95, 47.01),
    pub_md_base = c(0.68, 1.35, 2.03, 2.70, 3.38, 3.92, 4.73),
    pub_ld_salt = c(8.10, 16.20, 24.30, 32.40, 40.50, 46.98, 56.70),
    pub_md_salt = c(0.82, 1.63, 2.45, 3.26, 4.08, 4.73, 5.71)
  )

table2 |>
  transmute(
    target,
    ld_base = round(ld_base, 2), pub_ld_base,
    md_base = round(md_base, 2), pub_md_base,
    ld_salt = round(ld_salt, 2), pub_ld_salt,
    md_salt = round(md_salt, 2), pub_md_salt
  ) |>
  rename(
    "Target (ug/mL)"          = target,
    "Loading, base (model)"   = ld_base,
    "Loading, base (paper)"   = pub_ld_base,
    "Maint., base (model)"    = md_base,
    "Maint., base (paper)"    = pub_md_base,
    "Loading, salt (model)"   = ld_salt,
    "Loading, salt (paper)"   = pub_ld_salt,
    "Maint., salt (model)"    = md_salt,
    "Maint., salt (paper)"    = pub_md_salt
  ) |>
  knitr::kable(
    caption = paste(
      "Kuroda 2024 Table 2 recomputed from the packaged parameters.",
      "Loading doses in mg/kg, maintenance rates in mg/kg/h."
    )
  )

table2_dev <- with(table2, c(
  100 * (ld_base - pub_ld_base) / pub_ld_base,
  100 * (md_base - pub_md_base) / pub_md_base,
  100 * (ld_salt - pub_ld_salt) / pub_ld_salt,
  100 * (md_salt - pub_md_salt) / pub_md_salt
))
# Deterministic arithmetic; the residual difference is entirely the rounding of
# F to three digits (36.4% vs the ~36.48% the paper's own table implies:
# 4.90 / 26.86 * 2 = 0.3648). Realised max 1.02%.
stopifnot(max(abs(table2_dev)) < 2)
round(range(table2_dev), 3)
```

All fourteen published values reproduce to better than 1.1%. The residual is
traceable: the paper's table is internally consistent with `Vss/F = 13.431` and
`CL/F = 1.3499`, implying `F = 0.3648` rather than the 0.364 printed in Table 1.

# Virtual cohort

Original observed concentrations are not public. The cohorts below approximate
the published demographics. Because plasma concentration is invariant to body
weight in this model (see the assertion in the next chunk), the weight
distribution matters only for reporting the absolute doses in mg.

Cohorts are 150 horses per arm, well under the 200-per-arm cap. Kuroda 2024
used 5,000 virtual horses; 150 is ample to place a median and adequate for the
10th and 90th percentiles quoted below, whose Monte Carlo error is discussed
where they are compared.

```{r cohort}
# rxSetSeed() fixes rxode2's stream per solver thread, not across thread
# counts, so a CI runner with a different core count draws a different cohort.
# Every assertion on a cohort-derived quantity below is written to hold for any
# cohort this model can produce (see the bounds and their comments).
rxode2::rxSetSeed(20240607)
set.seed(20240607)

N_ARM <- 150L

# Doses, in mg/kg of quinidine sulfate dihydrate as the paper prescribes them,
# converted to QND base for the model.
DOSE <- list(
  iv_base          = 5 / SALT_HCL_MONOHYDRATE,       # 5 mg/kg HCl monohydrate -> 4.28
  po_base          = as_base_sulfate(20),            # 20 mg/kg salt          -> 16.58
  classical_base   = as_base_sulfate(22),            # 22 mg/kg salt          -> 18.24
  ldA_base         = as_base_sulfate(30),   mdA_base = as_base_sulfate(6.5),
  ldB_base         = as_base_sulfate(45),   mdB_base = as_base_sulfate(9),
  d1_ld_base       = as_base_sulfate(15),   d1_md_base = as_base_sulfate(3),
  d2_ld_base       = as_base_sulfate(30),   d2_md_base = as_base_sulfate(6.5),
  d3_ld_base       = as_base_sulfate(40),   d3_md_base = as_base_sulfate(9)
)
round(unlist(DOSE), 3)

# Build one arm. `id_offset` keeps subject IDs disjoint across arms; rxSolve
# treats id as the subject key and silently merges duplicates.
make_arm <- function(dose_times, dose_base_per_kg, arm, id_offset,
                     obs_times, route_iv = 0L, infusion_h = NA_real_) {
  stopifnot(length(dose_times) == length(dose_base_per_kg))
  wt <- runif(N_ARM, 430, 563)
  ids <- id_offset + seq_len(N_ARM)
  doses <- do.call(rbind, lapply(seq_along(dose_times), function(k) {
    data.frame(
      id = ids, time = dose_times[k], amt = dose_base_per_kg[k] * wt, evid = 1L,
      cmt = if (route_iv == 1L) "central" else "depot",
      dur = infusion_h, WT = wt, ROUTE_IV = route_iv, arm = arm
    )
  }))
  obs <- data.frame(
    id = rep(ids, each = length(obs_times)),
    time = rep(obs_times, times = N_ARM),
    amt = NA_real_, evid = 0L,
    # Observation rows point at the ODE state `central`, never at the
    # algebraic observable `Cc`; rxode2 returns Cc as a column regardless.
    cmt = "central", dur = NA_real_,
    WT = rep(wt, each = length(obs_times)),
    ROUTE_IV = route_iv, arm = arm
  )
  rbind(doses, obs)
}
```

# Single-dose arms: intravenous and oral

These two arms reproduce the healthy-horse crossover of Kuroda 2024 Figures 1
and 2, and they are what the NCA section below is computed from. The grid is
dense over the absorption and distribution phases (the alpha half-life is only
3.4 minutes) and runs to 72 h so the 7.76 h terminal phase is well resolved.

```{r single-dose}
grid_sd <- sort(unique(c(
  seq(0, 2, by = 0.1), seq(2, 12, by = 0.25), seq(12, 72, by = 1)
)))

events_sd <- bind_rows(
  make_arm(0, DOSE$iv_base, "IV 4.28 mg/kg base (5-min infusion)",
           0L, grid_sd, route_iv = 1L, infusion_h = 5 / 60),
  make_arm(0, DOSE$po_base, "PO 16.58 mg/kg base (nasogastric)",
           1000L, grid_sd, route_iv = 0L)
)
stopifnot(!anyDuplicated(unique(events_sd[, c("id", "time", "evid")])))

sim_sd <- rxode2::rxSolve(mod, events_sd, keep = c("arm", "WT"),
                          returnType = "data.frame")

# Typical-value (zeroRe) counterpart for the deterministic NCA checks.
events_typ <- events_sd |>
  filter(id %in% c(1L, 1001L)) |>
  mutate(amt = ifelse(evid == 1L, amt / WT * 490, amt), WT = 490)
sim_typ <- rxode2::rxSolve(rxode2::zeroRe(mod), events_typ,
                           keep = c("arm"), returnType = "data.frame")

# Body weight cancels out of the concentration: all volumes and clearances
# scale linearly with WT and the dose is prescribed per kilogram, so
# amt / (per-kg volume * WT) is independent of WT. Built explicitly here (one
# subject per weight, disjoint ids) rather than via make_arm, which draws its
# own weights.
wt_one <- function(id, wt) {
  rbind(
    data.frame(id = id, time = 0, amt = DOSE$po_base * wt, evid = 1L,
               cmt = "depot", dur = NA_real_, WT = wt, ROUTE_IV = 0L),
    data.frame(id = id, time = grid_sd, amt = NA_real_, evid = 0L,
               cmt = "central", dur = NA_real_, WT = wt, ROUTE_IV = 0L)
  )
}
wt_check <- rxode2::rxSolve(
  rxode2::zeroRe(mod),
  rbind(wt_one(9001L, 430), wt_one(9002L, 563)),
  returnType = "data.frame"
)
wt_cmax <- tapply(wt_check$Cc, wt_check$id, max, na.rm = TRUE)
wt_rel <- abs(diff(range(wt_cmax))) / mean(wt_cmax)
cat("Cmax at 430 kg and 563 kg agree to", signif(wt_rel, 3), "relative\n")
# Analytically identical, and it comes back identical: the two solves agree to
# ~2e-16, i.e. machine precision. Deterministic, so the bound is tight.
stopifnot(wt_rel < 1e-8)
```

```{r figure-1-2}
# Replicates Figures 1 and 2 of Kuroda 2024: semilogarithmic disposition curves
# after single intravenous and single oral administration in healthy horses.
sim_sd |>
  filter(!is.na(Cc), Cc > 0, time > 0) |>
  group_by(arm, time) |>
  summarise(
    p10 = quantile(Cc, 0.10), p50 = quantile(Cc, 0.50),
    p90 = quantile(Cc, 0.90), .groups = "drop"
  ) |>
  ggplot(aes(time, p50)) +
  geom_ribbon(aes(ymin = p10, ymax = p90), alpha = 0.25, fill = "steelblue") +
  geom_line(linewidth = 0.7) +
  facet_wrap(~arm) +
  scale_x_continuous(limits = c(0, 24)) +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Plasma quinidine, QND base (ug/mL)",
    title = "Single-dose disposition, median with 10th-90th percentile band",
    caption = "Replicates Figures 1 and 2 of Kuroda 2024 (n = 150 per arm)."
  )
```

# PKNCA validation

Kuroda 2024 reports no NCA table, but four of its Table 1 rows are quantities
NCA computes directly from the intravenous arm: plasma clearance, the terminal
half-life, the steady-state volume of distribution, and the mean residence
time. Running PKNCA on the typical-value intravenous profile therefore checks
the packaged parameters against the published values through an independent
route -- numerical integration of the solved ODE rather than the algebra of the
previous section.

```{r pknca-iv}
conc_iv <- sim_typ |>
  filter(arm == "IV 4.28 mg/kg base (5-min infusion)", !is.na(Cc)) |>
  select(id, time, Cc) |>
  mutate(treatment = "IV")

# Guarantee a time-zero row. Filter on !is.na(Cc) only -- a `time > 0` or
# `Cc > 0` filter would drop it and trigger PKNCA's "AUC range starting (0)
# before the first measurement" warning on every subject.
conc_iv <- bind_rows(
  conc_iv,
  conc_iv |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, time)

dose_iv <- events_typ |>
  filter(evid == 1L, arm == "IV 4.28 mg/kg base (5-min infusion)") |>
  select(id, time, amt) |>
  mutate(treatment = "IV")

nca_iv <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(as.data.frame(conc_iv), Cc ~ time | treatment + id,
                   concu = "ug/mL", timeu = "h"),
  PKNCA::PKNCAdose(as.data.frame(dose_iv), amt ~ time | treatment + id,
                   doseu = "mg", route = "intravascular", duration = 5 / 60),
  intervals = data.frame(
    start = 0, end = Inf,
    cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE,
    cl.obs = TRUE, vss.obs = TRUE, mrt.obs = TRUE
  )
))

iv_pp <- as.data.frame(nca_iv) |>
  select(PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

# PKNCA returns CL in L/h and Vss in L for a 490 kg horse; the paper reports
# both per kilogram.
WT_TYPICAL <- 490
iv_compare <- tibble::tibble(
  parameter = c("CL", "Vss", "MRT", "Terminal half-life"),
  units     = c("L/kg/h", "L/kg", "h", "h"),
  nca       = c(iv_pp$cl.obs / WT_TYPICAL, iv_pp$vss.obs / WT_TYPICAL,
                iv_pp$mrt.obs, iv_pp$half.life),
  published = c(0.49, 4.90, 10.12, 7.76)
) |>
  mutate(pct_diff = 100 * (nca - published) / published)

iv_compare |>
  mutate(across(c(nca, published, pct_diff), \(x) round(x, 3))) |>
  rename("Parameter" = parameter, "Units" = units,
         "PKNCA on the typical-value IV profile" = nca,
         "Kuroda 2024 Table 1" = published, "% difference" = pct_diff) |>
  knitr::kable(caption = "Noncompartmental analysis of the intravenous arm vs Kuroda 2024 Table 1.")

# Deterministic: one typical-value profile, so the only error is numerical
# integration on a finite grid. CL is Dose/AUCinf and is the tightest.
stopifnot(
  abs(iv_compare$pct_diff[iv_compare$parameter == "CL"]) < 2,
  abs(iv_compare$pct_diff[iv_compare$parameter == "Terminal half-life"]) < 3,
  abs(iv_compare$pct_diff[iv_compare$parameter == "Vss"]) < 5,
  abs(iv_compare$pct_diff[iv_compare$parameter == "MRT"]) < 5
)
```

## Recovering the bioavailability factor

The ratio of dose-normalised exposures between the oral and intravenous arms is
an independent estimate of `F`, and it exercises the logit-scale encoding of the
bioavailability parameter end to end.

```{r pknca-f}
conc_po <- sim_typ |>
  filter(arm == "PO 16.58 mg/kg base (nasogastric)", !is.na(Cc)) |>
  select(id, time, Cc) |>
  mutate(treatment = "PO")
conc_po <- bind_rows(
  conc_po,
  conc_po |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, time)

dose_po <- events_typ |>
  filter(evid == 1L, arm == "PO 16.58 mg/kg base (nasogastric)") |>
  select(id, time, amt) |>
  mutate(treatment = "PO")

nca_po <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(as.data.frame(conc_po), Cc ~ time | treatment + id,
                   concu = "ug/mL", timeu = "h"),
  PKNCA::PKNCAdose(as.data.frame(dose_po), amt ~ time | treatment + id,
                   doseu = "mg", route = "extravascular"),
  intervals = data.frame(
    start = 0, end = Inf,
    cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
  )
))

po_pp <- as.data.frame(nca_po) |>
  select(PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

f_recovered <-
  (po_pp$aucinf.obs / (DOSE$po_base * WT_TYPICAL)) /
  (iv_pp$aucinf.obs / (DOSE$iv_base * WT_TYPICAL))

cat("F recovered from the AUCinf ratio:", round(f_recovered, 4),
    " | Kuroda 2024 Table 1: 0.364\n")
cat("Oral Tmax:", round(po_pp$tmax, 3), "h  (absorption half-life",
    round(log(2) / TV$ka, 2), "h)\n")

# Deterministic identity, limited only by the trapezoidal grid.
stopifnot(abs(100 * (f_recovered - TV$fdepot) / TV$fdepot) < 2)
```

`F` comes back at essentially the published 36.4%, which confirms that placing
the random effect on the logit scale (and mapping back with `expit()`) leaves
the typical value where Table 1 puts it.

# Monte Carlo simulation of the dosing regimens

This is the paper's principal output. Kuroda 2024 simulates the classical
regimen and three proposed alternatives, and reports selected percentiles of the
plasma concentration reached. Five arms are reproduced here.

An important point of interpretation. Kuroda 2024 derives the maintenance rates
from `C_target * CL / F`, an equation whose solution is the **average**
concentration over a dosing interval at steady state, not a peak or a trough.
And the published 50th percentiles for the two Figure 8 regimens are exactly
what that equation returns:

```{r design-identity}
design_identity <- tibble::tibble(
  regimen = c("Loading 30 mg/kg + 6.5 mg/kg q2h", "Loading 45 mg/kg + 9 mg/kg q2h"),
  md_salt_per_dose = c(6.5, 9.0),
  tau_h = c(2, 2)
) |>
  mutate(
    rate_base = as_base_sulfate(md_salt_per_dose) / tau_h,
    c_avg_ss  = rate_base * TV$fdepot / TV$cl,
    published_p50 = c(2.0, 2.8),
    pct_diff = 100 * (c_avg_ss - published_p50) / published_p50
  )

design_identity |>
  mutate(across(c(rate_base, c_avg_ss, pct_diff), \(x) round(x, 3))) |>
  rename("Regimen" = regimen, "Maint. dose (mg/kg salt)" = md_salt_per_dose,
         "tau (h)" = tau_h, "Rate (mg/kg/h base)" = rate_base,
         "C_avg,ss = rate * F / CL" = c_avg_ss,
         "Kuroda 2024 50th pct" = published_p50, "% difference" = pct_diff) |>
  knitr::kable(caption = "Kuroda 2024 Figure 8 median concentrations vs the paper's own design equation.")

# Deterministic; reproduces the published medians to better than 1.5%.
stopifnot(max(abs(design_identity$pct_diff)) < 2)
```

So the statistic to compare against is the average concentration over a dosing
interval, which PKNCA computes as `cav`. The regimen arms below are therefore
observed on a fine grid across the final dosing interval of interest, and
compared on `cav`.

```{r regimens}
grid_24 <- sort(unique(c(seq(0, 22, by = 1), seq(22, 24, by = 0.1))))
grid_12 <- sort(unique(c(seq(0, 6, by = 1), seq(6, 12, by = 0.1))))
grid_72 <- sort(unique(c(seq(0, 72, by = 1), seq(6, 8, by = 0.1),
                         seq(30, 32, by = 0.1), seq(54, 56, by = 0.1))))

events_reg <- bind_rows(
  # Figure 8, regimen A: loading 30 mg/kg then 6.5 mg/kg q2h for 24 h.
  make_arm(seq(0, 22, by = 2),
           c(DOSE$ldA_base, rep(DOSE$mdA_base, 11)),
           "Loading 30 + 6.5 q2h", 2000L, grid_24),
  # Figure 8, regimen B: loading 45 mg/kg then 9 mg/kg q2h.
  make_arm(seq(0, 22, by = 2),
           c(DOSE$ldB_base, rep(DOSE$mdB_base, 11)),
           "Loading 45 + 9 q2h", 2200L, grid_24),
  # Figure 7: the classical 22 mg/kg regimen, q2h to four doses and q6h twice.
  make_arm(c(0, 2, 4, 6), rep(DOSE$classical_base, 4),
           "Classical 22 q2h x4", 2400L, grid_12),
  make_arm(c(0, 6), rep(DOSE$classical_base, 2),
           "Classical 22 q6h x2", 2600L, grid_12),
  # Figure 9: three days of progressively increasing daytime dosing.
  make_arm(c(0, 2, 4, 6, 24, 26, 28, 30, 48, 50, 52, 54),
           c(DOSE$d1_ld_base, rep(DOSE$d1_md_base, 3),
             DOSE$d2_ld_base, rep(DOSE$d2_md_base, 3),
             DOSE$d3_ld_base, rep(DOSE$d3_md_base, 3)),
           "3-day progressive", 2800L, grid_72)
)
stopifnot(!anyDuplicated(unique(events_reg[, c("id", "time", "evid")])))

sim_reg <- rxode2::rxSolve(mod, events_reg, keep = c("arm"),
                           returnType = "data.frame")
stopifnot(all(sim_reg$Cc[!is.na(sim_reg$Cc)] >= 0))

# Average concentration over an arbitrary window, via PKNCA's cav.
interval_cav <- function(sim, arm_name, start, end) {
  cc <- sim |>
    filter(arm == arm_name, !is.na(Cc)) |>
    select(id, time, Cc) |>
    mutate(treatment = arm_name)
  dd <- events_reg |>
    filter(arm == arm_name, evid == 1L) |>
    select(id, time, amt) |>
    mutate(treatment = arm_name)
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(as.data.frame(cc), Cc ~ time | treatment + id,
                     concu = "ug/mL", timeu = "h"),
    PKNCA::PKNCAdose(as.data.frame(dd), amt ~ time | treatment + id,
                     doseu = "mg", route = "extravascular"),
    intervals = data.frame(start = start, end = end,
                           cav = TRUE, cmax = TRUE, cmin = TRUE)
  ))
  as.data.frame(res) |>
    filter(PPTESTCD %in% c("cav", "cmax", "cmin")) |>
    select(id, PPTESTCD, PPORRES) |>
    tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
    mutate(arm = arm_name)
}
```

## Figure 8: the two proposed 24-hour regimens

```{r figure-8}
# Replicates Figure 8 of Kuroda 2024: percentiles of plasma quinidine at 24 h
# under the two proposed loading-plus-maintenance regimens. Compared on the
# average concentration over the final (22-24 h) dosing interval.
fig8 <- bind_rows(
  interval_cav(sim_reg, "Loading 30 + 6.5 q2h", 22, 24),
  interval_cav(sim_reg, "Loading 45 + 9 q2h", 22, 24)
)

fig8_summary <- fig8 |>
  group_by(arm) |>
  summarise(
    p10 = quantile(cav, 0.10), p50 = quantile(cav, 0.50),
    p90 = quantile(cav, 0.90), .groups = "drop"
  ) |>
  mutate(
    pub_p10 = c(1.4, 1.9), pub_p50 = c(2.0, 2.8), pub_p90 = c(2.7, 3.8),
    pct_p50 = 100 * (p50 - pub_p50) / pub_p50
  )

fig8_summary |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  rename("Regimen" = arm, "10th" = p10, "50th" = p50, "90th" = p90,
         "10th (paper)" = pub_p10, "50th (paper)" = pub_p50,
         "90th (paper)" = pub_p90, "% diff, median" = pct_p50) |>
  knitr::kable(caption = "Figure 8 of Kuroda 2024: simulated vs published percentiles of Cav over the final dosing interval (ug/mL).")

# Cohort-derived, so the bound must hold for any cohort this model can draw.
# Realised median deviations of -3.7% to -8.0% across 1, 2, 4, 8 and 16 solver
# threads (rxode2's RNG streams are partitioned per thread, so a CI runner
# draws a different cohort). 20% leaves better than 2x headroom over that
# range while a mis-transcribed clearance, dose or salt factor moves the
# median by tens of percent and still breaks this.
stopifnot(max(abs(fig8_summary$pct_p50)) < 20)

sim_reg |>
  filter(arm %in% c("Loading 30 + 6.5 q2h", "Loading 45 + 9 q2h"), !is.na(Cc)) |>
  group_by(arm, time) |>
  summarise(p10 = quantile(Cc, 0.10), p50 = quantile(Cc, 0.50),
            p90 = quantile(Cc, 0.90), .groups = "drop") |>
  ggplot(aes(time, p50)) +
  geom_ribbon(aes(ymin = p10, ymax = p90), alpha = 0.25, fill = "steelblue") +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 2.0, linetype = "dashed", colour = "darkgreen") +
  geom_hline(yintercept = 3.8, linetype = "dotted", colour = "firebrick") +
  facet_wrap(~arm) +
  labs(
    x = "Time (h)", y = "Plasma quinidine, QND base (ug/mL)",
    title = "Figure 8: proposed loading-plus-maintenance regimens",
    caption = paste(
      "Replicates Figure 8 of Kuroda 2024. Dashed line 2.0 ug/mL (median",
      "therapeutic), dotted line 3.8 ug/mL (median toxic)."
    )
  )
```

## Figure 9: three days of progressively increasing dosing

The regimen the paper actually recommends. Each day gives a loading dose
followed, two hours later, by three maintenance doses at two-hour intervals,
with the doses escalating day over day; overnight the concentration is allowed
to fall.

```{r figure-9}
# Replicates Figure 9 of Kuroda 2024. The comparison statistic is the average
# concentration over the final dosing interval of each day's block.
fig9 <- bind_rows(lapply(1:3, function(d) {
  start <- c(6, 30, 54)[d]
  interval_cav(sim_reg, "3-day progressive", start, start + 2) |>
    mutate(day = d)
}))

fig9_summary <- fig9 |>
  group_by(day) |>
  summarise(p10 = quantile(cav, 0.10), p50 = quantile(cav, 0.50),
            p90 = quantile(cav, 0.90), .groups = "drop") |>
  mutate(
    pub_p10 = c(0.6, 1.3, 1.9), pub_p50 = c(0.8, 1.8, 2.6),
    pub_p90 = c(1.1, 2.7, 3.8),
    pct_p50 = 100 * (p50 - pub_p50) / pub_p50
  )

fig9_summary |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  rename("Day" = day, "10th" = p10, "50th" = p50, "90th" = p90,
         "10th (paper)" = pub_p10, "50th (paper)" = pub_p50,
         "90th (paper)" = pub_p90, "% diff, median" = pct_p50) |>
  knitr::kable(caption = "Figure 9 of Kuroda 2024: simulated vs published percentiles of Cav over each day's final dosing interval (ug/mL).")

# Realised median deviations of +0.3% to -8.3% across 1, 2, 4, 8 and 16 solver
# threads. Same reasoning and same bound as Figure 8.
stopifnot(max(abs(fig9_summary$pct_p50)) < 20)

sim_reg |>
  filter(arm == "3-day progressive", !is.na(Cc)) |>
  group_by(time) |>
  summarise(p10 = quantile(Cc, 0.10), p50 = quantile(Cc, 0.50),
            p90 = quantile(Cc, 0.90), .groups = "drop") |>
  ggplot(aes(time, p50)) +
  geom_ribbon(aes(ymin = p10, ymax = p90), alpha = 0.25, fill = "steelblue") +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 2.0, linetype = "dashed", colour = "darkgreen") +
  geom_hline(yintercept = 3.8, linetype = "dotted", colour = "firebrick") +
  scale_x_continuous(breaks = seq(0, 72, by = 12)) +
  labs(
    x = "Time (h)", y = "Plasma quinidine, QND base (ug/mL)",
    title = "Figure 9: three-day progressive regimen",
    caption = paste(
      "Replicates Figure 9 of Kuroda 2024. Dashed line 2.0 ug/mL,",
      "dotted line 3.8 ug/mL."
    )
  )
```

## Figure 7: the classical 22 mg/kg regimen

Kuroda 2024 simulates the textbook regimen in order to argue against it: after
four q2h doses, 10% of horses exceed 4.8 ug/mL, well above the 3.8 ug/mL median
toxic concentration established in the same paper.

Unlike Figures 8 and 9, the statistic behind the Figure 7 numbers is not stated,
and it cannot be the interval average -- for the q2h arm the paper's own design
equation gives an average steady-state concentration of 6.8 ug/mL, far above the
quoted percentiles, so the figure must be reporting an earlier, non-steady-state
point. The peak within the interval following the fourth dose is the reading
that best matches the published wording ("after the 4th q2h administration"),
and it is used below. **The q6h arm does not reproduce on any statistic tried
and is recorded as a known deviation rather than gated.**

```{r figure-7}
fig7 <- bind_rows(
  interval_cav(sim_reg, "Classical 22 q2h x4", 6, 8),
  interval_cav(sim_reg, "Classical 22 q6h x2", 6, 12)
)

fig7_summary <- fig7 |>
  group_by(arm) |>
  summarise(
    cmax_p10 = quantile(cmax, 0.10), cmax_p50 = quantile(cmax, 0.50),
    cmax_p90 = quantile(cmax, 0.90), .groups = "drop"
  ) |>
  mutate(
    pub_p10 = c(2.5, 1.9), pub_p90 = c(4.8, 3.9),
    pct_p10 = 100 * (cmax_p10 - pub_p10) / pub_p10,
    pct_p90 = 100 * (cmax_p90 - pub_p90) / pub_p90,
    deviation = c(FALSE, TRUE)
  )

fig7_summary |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  rename("Regimen" = arm, "Cmax 10th" = cmax_p10, "Cmax 50th" = cmax_p50,
         "Cmax 90th" = cmax_p90, "10th (paper)" = pub_p10,
         "90th (paper)" = pub_p90, "% diff, 10th" = pct_p10,
         "% diff, 90th" = pct_p90, "Known deviation" = deviation) |>
  knitr::kable(caption = "Figure 7 of Kuroda 2024: simulated vs published percentiles of the peak concentration in the interval after the final dose (ug/mL).")

# Only the q2h arm is gated. These are tail percentiles of a 150-horse cohort,
# so they move more than a median does: realised deviations spanned -14.7% to
# +18.0% across 1, 2, 4, 8 and 16 solver threads. 30% keeps headroom over that
# range without becoming unfalsifiable -- the published percentiles are 2.5 and
# 4.8 ug/mL, so 30% is still a 0.75-1.4 ug/mL window.
gated7 <- fig7_summary |> filter(!deviation)
stopifnot(max(abs(c(gated7$pct_p10, gated7$pct_p90))) < 30)

# The paper's qualitative conclusion -- that the classical q2h regimen pushes a
# tenth of the population past the 3.8 ug/mL median toxic concentration -- is
# the load-bearing claim of Figure 7, and it does reproduce.
q2h_p90 <- fig7_summary$cmax_p90[fig7_summary$arm == "Classical 22 q2h x4"]
cat("Classical q2h, 90th percentile of peak concentration:",
    round(q2h_p90, 2), "ug/mL vs the 3.8 ug/mL median toxic concentration\n")
stopifnot(q2h_p90 > 3.8)
```

## Probability of target attainment in the therapeutic window

Kuroda 2024 states that targeting 2.9 ug/mL "was expected to include the largest
portion of the population in the therapeutic range (a PTA of 80% was achieved
between 2.0 and 3.8 ug/mL)". Regimen B is the 2.9 ug/mL target rounded for
prescribing.

```{r pta}
pta <- fig8 |>
  group_by(arm) |>
  summarise(
    pta_pct = 100 * mean(cav >= 2.0 & cav <= 3.8),
    below_pct = 100 * mean(cav < 2.0),
    above_pct = 100 * mean(cav > 3.8),
    .groups = "drop"
  )

pta |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  rename("Regimen" = arm, "PTA 2.0-3.8 ug/mL (%)" = pta_pct,
         "Below window (%)" = below_pct, "Above window (%)" = above_pct) |>
  knitr::kable(caption = "Probability of target attainment within the 2.0-3.8 ug/mL therapeutic window.")

# Realised 71.3% to 76.0% across 1, 2, 4, 8 and 16 solver threads against the
# paper's quoted 80%. A two-sided band around that, wide enough for the
# cohort-to-cohort spread but narrow enough that a wrong clearance or salt
# factor (which would push the whole distribution out of the window) breaks it.
pta_b <- pta$pta_pct[pta$arm == "Loading 45 + 9 q2h"]
stopifnot(pta_b > 60, pta_b < 95)
cat("Regimen B PTA:", round(pta_b, 1), "% vs the paper's quoted 80%\n")
```

# Assumptions and deviations

## Errata in the source

* **Additive residual error units.** Table 1 labels `Stdev0` and `Stdev1`
  "ug/L", and section 2.5 says the additive sigma was "reported as its standard
  deviation noted with the same units as plasma concentration (ug/L)". Both are
  typographical: the assay's lower limit of quantitation is 0.03 ug/mL and every
  concentration in the paper is in ug/mL, so an additive SD of 0.0338 ug/L
  would be a thousandfold below the quantification limit. Read as ug/mL the two
  values (0.0338 and 0.0479) land essentially at the limit of quantitation,
  which is what a well-behaved additive term does. The model carries them as
  ug/mL. The sibling model `Kuroda_2023_cephalothin.R`, from the same group,
  carries the identical typo and resolves it the same way.

* **The QND base equivalent of the 6.5 mg/kg maintenance dose.** The Results
  section gives the Figure 8 regimen A maintenance dose as "6.5 mg/kg
  (4.97 mg/kg as QND base)". Every other salt-to-base conversion in the paper is
  exactly `salt / 1.206`, which gives 5.39 mg/kg, not 4.97; the printed 4.97
  corresponds to a 6.0 mg/kg salt dose. Two independent lines of evidence show
  6.5 mg/kg salt is the intended dose and 4.97 is the stale number: the salt
  figure 6.5 is repeated consistently in the Abstract, Results and Discussion,
  and the paper's own maintenance-rate equation for the 2.0 ug/mL target gives
  2.70 mg/kg/h base (Table 2), which over a two-hour interval is 5.40 mg/kg base
  and 6.52 mg/kg salt. This vignette uses 6.5 mg/kg salt = 5.39 mg/kg base.
  Nothing in the model file depends on the erratum.

* **Subject count.** Section 2.1 and the Abstract both give 29 horses
  (10 healthy plus 19 with atrial fibrillation), and the Abstract says "the data
  from 29 horses were modeled". The Table 1 caption instead reads "in
  27 horses". The paper never reconciles the two or names an exclusion. The
  model metadata records 29, following the two concordant sources.

* **Rounding of F.** Table 1 prints `F = 36.4%`, but Table 2 is internally
  consistent with `Vss/F = 13.431` and `CL/F = 1.3499`, both of which imply
  `F = 0.3648`. The model carries the printed 0.364; this is the sole source of
  the sub-1.1% residual in the Table 2 reproduction above.

## Modelling assumptions

* **OMEGA off-diagonals are not available.** Section 2.5 states that "a full
  OMEGA matrix was used to determine the random components of the model", but
  Table 1 reports only the diagonal (the BSV% column). No correlation,
  covariance or OMEGA block appears anywhere in the paper or its figures. Per
  the standing policy for unreported variance components, only the eight
  reported variances are packaged and the off-diagonals are left at zero rather
  than invented. The between-subject spread simulated here is therefore not
  exactly the paper's, which is one contributor to the residual differences in
  the Figure 7-9 percentile tables. A user who obtains the full matrix can
  supply it directly.

* **BSV% to variance conversion.** Section 2.5 Equation 2 defines the reported
  BSV% as `100 * sqrt(exp(omega^2) - 1)`, so each variance is recovered as
  `omega^2 = log(1 + (BSV%/100)^2)`. This is the exponential-model convention,
  and the equation is printed explicitly, so no interpretation was needed.

* **Bioavailability on the logit scale.** The Table 1 footnote records that "for
  F, an ilogit transformation was used". `F` is therefore held as
  `logitfdepot = qlogis(0.364)` with its random effect added on the logit scale
  and mapped back with `expit()`. Encoding it as `exp(lfdepot + eta)` instead
  would place roughly 0.1% of simulated horses above 100% bioavailability, which
  is exactly what the paper's transformation exists to prevent. The AUC-ratio
  check above confirms the typical value is unaffected.

* **Route-specific residual error.** Table 1 estimates a separate proportional
  and additive pair for the intravenous and the oral datasets. The model
  reproduces both and selects between them with the canonical `ROUTE_IV`
  indicator, following `Ahmed_2015_topiramate.R` and
  `Fanta_2007_ciclosporin.R`. Set `ROUTE_IV = 1` and dose into `central` for
  intravenous administration, `ROUTE_IV = 0` and dose into `depot` for oral.

* **No covariates.** None of condition (healthy or AF), age, body weight or sex
  reached the paper's BIC threshold, so no covariate effect is carried on any
  structural parameter. All four are recorded in the model's
  `covariatesDataExcluded` metadata to preserve the provenance of the screen.
  Body weight is the exception: it appears in `covariateData` because the
  per-kilogram parameterisation of Table 1 requires it as a linear scaler, which
  is the base model and not a covariate effect.

* **Body weight distribution.** Drawn uniformly over 430-563 kg, the union of
  the two cohorts' reported ranges. The choice is immaterial to the
  concentrations, as the assertion in the single-dose chunk demonstrates: weight
  cancels exactly.

* **Figure 7's q6h arm is a known deviation.** Against the published 10th and
  90th percentiles of 1.9 and 3.9 ug/mL, the model's interval-peak percentiles
  come out roughly 34% and 17% low, and the interval-average statistic is lower
  still (about half the published values) -- while the q2h arm of the same
  figure reproduces within about 15-18% on the same statistic. Two features of
  the paper make this hard to pin down: it does not state what statistic
  Figure 7 plots, and it does not say how long the q6h simulation was run
  (section 2.5 describes the classical regimen as "administered twice every
  6 h", but percentiles near 3.9 ug/mL would require dosing continued closer to
  steady state, where the paper's own design equation puts the average
  concentration at 2.26 ug/mL). Rather than widen the gate until the
  disagreement disappeared, the row is left in the rendered table, flagged in
  the `Known deviation` column, and excluded from the assertion.

* **Cohort size.** 150 horses per arm against the paper's 5,000. Percentiles at
  the 10th and 90th carry visible Monte Carlo error at this size, which is
  reflected in the width of the bounds asserted on them; the medians are well
  determined.
