---
title: "Bedaquiline (Kurosawa 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bedaquiline (Kurosawa 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Kurosawa_2021_bedaquiline"))$reference`
* Description: `r rxode2::rxode(readModelDb("Kurosawa_2021_bedaquiline"))$description`
* Article: <https://doi.org/10.1002/jcph.1887> (open access, PMC8518967)
* Previously developed model: McLeay et al. 2014,
  <https://doi.org/10.1128/AAC.01418-13>

Kurosawa 2021 ran a phase 1 crossover study (NCT03800550) of a single 100 mg
bedaquiline tablet given alone (treatment A) or on day 5 of clarithromycin
500 mg every 12 hours (treatment B). The population PK analysis took the
four-compartment bedaquiline model of McLeay 2014, confirmed that it described
the monotherapy data, and then estimated a single new parameter: the effect of
clarithromycin on bedaquiline apparent clearance,

$$CL/F = CL_{pop} \cdot (1 + \theta)^{CLR_i}, \qquad \theta = -0.37,$$

a 37% reduction. Every other parameter was carried unchanged from McLeay 2014
(Kurosawa 2021 Supplemental Table S1, whose "updated model" column is blank for
everything except the clarithromycin effect), and is therefore `fixed()` in the
packaged model. The updated model was then used to simulate 48-week bedaquiline
regimens with clarithromycin for pulmonary nontuberculous mycobacterial disease.

# Population

The updated model was estimated on 16 healthy White adults in Belgium (9 women,
7 men), median age 43 years (24-55), median weight 65.65 kg (56.8-100.0) and
median BMI 22.44 kg/m^2 (18.9-29.5) (Kurosawa 2021 Table 1). The fixed
parameters come from McLeay 2014: 5,222 concentrations from 480 subjects
(111 healthy volunteers, 44 patients with drug-sensitive TB and 325 patients
with MDR-TB) in nine phase I/II studies.

```{r population}
str(readModelDb("Kurosawa_2021_bedaquiline")()$population)
```

# Source trace

Every value is also traced in a comment beside its `ini()` line in
`inst/modeldb/specificDrugs/Kurosawa_2021_bedaquiline.R`. "S1" is Kurosawa 2021
Supplemental Table S1 (which reprints the McLeay 2014 Table 3 estimates in its
"previously developed model" column).

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL/F, reference individual) | 2.78 L/h, fixed | S1; McLeay 2014 Table 3 |
| `lvc` (Vc/F) | 164 L, fixed | S1 |
| `lq`, `lvp` (CLp1/F, Vp1/F) | 11.8 L/h, 178 L, fixed | S1 |
| `lq2`, `lvp2` (CLp2/F, Vp2/F) | 8.03 L/h, 3010 L, fixed | S1 |
| `lq3`, `lvp3` (CLp3/F, Vp3/F) | 3.58 L/h, 7350 L, fixed | S1 |
| `lka` (input compartment to central) | 1000 1/h, fixed | McLeay 2014 Results and Fig. 1 |
| `logitfrel` (FR1, fraction into the depot arm) | 58.5%, fixed | S1; logit constraint, McLeay 2014 Results |
| `ld1`, `ld2` (DUR1, DUR2) | 2.22 h, 1.48 h, fixed | S1 'D1', 'D2' |
| `ltlag`, `ltlag_tablet` (Tlag) | 0.541 h solution, 0.917 h tablet, fixed | S1 'ALAG1 solution', 'ALAG1 tablet' |
| `ltlag_add` (Tlag,add) | 1.48 h, fixed | S1 'TLAG' |
| lag of the central arm | Tlag + Tlag,add | McLeay 2014 Fig. 1 |
| `lfdepot` (F, studies C208 / C209) | 1, fixed reference | McLeay 2014 Results |
| `e_study_bdq_cde102_c104_f` | 1.51, fixed | S1 |
| `e_study_other_f` | 2.03, fixed | S1 'Other studies on F' |
| `e_race_black_cl` | +52.0%, fixed | S1 |
| `e_nonmdr_cl` (healthy volunteers / DS-TB) | +37.5%, fixed | S1 |
| `e_sexf_vc` | -15.7%, fixed | S1 |
| `e_conmed_clarithromycin_cl` | -0.37 (RSE 11%), estimated | S1 'Effect CLR on CL/F'; Results |
| Covariate form `(1 + theta)^cov` | n/a | Kurosawa 2021 Methods equation; McLeay 2014 equation 5 |
| `etalcl`, `etalvc` | 0.504^2, 0.391^2, correlation 0.407 | S1 'BSV' column and 'Correlation CL/Vc' |
| `etalogitfrel` | 1.13^2 (logit scale) | S1 'BSV' FR1 = 113 |
| `etalfdepot` | 0.396^2 | S1 'Between-subject variability on F' |
| `expSd`, `expSdC208C209` | 0.206, 0.277 | S1 'RUV' rows; log-transform-both-sides, McLeay 2014 equation 2 |
| `d/dt()` system | n/a | McLeay 2014 Fig. 1 (four compartments, dual zero-order input) |

# Dual zero-order absorption

Each oral dose is split: a fraction FR1 = 58.5% enters an input compartment as
a zero-order infusion over DUR1 = 2.22 h, starting after the lag Tlag, and is
passed to the central compartment at 1,000 1/h (so this arm is effectively
zero-order too); the remaining 41.5% enters the central compartment directly
as a zero-order infusion over DUR2 = 1.48 h, starting after Tlag + Tlag,add.
Every administration is therefore **two dose records at the same time**, one
to `depot` and one to `central`, each carrying the whole dose amount (the
`f()` terms perform the split) and its zero-order duration.

The durations are passed explicitly in a `dur` column. They are fixed
constants with no random effect, so this is numerically the same as
requesting the modelled durations with `rate = -2`, and it avoids a failure of
some rxode2 releases to solve two simultaneous modelled-duration doses into
different compartments.

```{r helpers}
mod <- readModelDb("Kurosawa_2021_bedaquiline")
ui <- rxode2::rxode2(mod)
modTyp <- rxode2::zeroRe(ui)

DUR1 <- 2.22 # S1 'D1', fixed
DUR2 <- 1.48 # S1 'D2', fixed

# One row per subject per dose time and arm; `id` and covariates are added
# by the caller.
doseRows <- function(times, amt) {
  dplyr::bind_rows(
    tibble(time = times, amt = amt, cmt = "depot", evid = 1L, dur = DUR1),
    tibble(time = times, amt = amt, cmt = "central", evid = 1L, dur = DUR2)
  )
}
# Observation rows target the ODE state `central`; rxode2 returns Cc.
obsRows <- function(times) {
  tibble(time = times, amt = NA_real_, cmt = "central", evid = 0L, dur = NA_real_)
}

# Covariates of the Kurosawa 2021 healthy volunteers: 100 mg tablet, not
# MDR-TB (so CL/F carries the +37.5% healthy-volunteer effect), a new study
# (so relative F is the 'other studies' 2.03), all White.
addHvCov <- function(d) {
  dplyr::mutate(d,
    RACE_BLACK = 0L, DIS_TB_MDR = 0L, STUDY_BDQ_C208_C209 = 0L,
    STUDY_BDQ_CDE102_C104 = 0L, FORM_TABLET = 1L
  )
}
```

```{r dual-peak, fig.cap = "Typical 100 mg tablet profile in a healthy male volunteer over the first 12 h, and the two input arms."}
evTyp <- dplyr::bind_rows(doseRows(0, 100), obsRows(seq(0, 12, by = 0.02))) |>
  dplyr::arrange(time, dplyr::desc(evid)) |>
  dplyr::mutate(id = 1L, SEXF = 0L, CONMED_CLARITHROMYCIN = 0L) |>
  addHvCov()
simTyp <- rxode2::rxSolve(modTyp, evTyp, returnType = "data.frame",
                          rtol = 1e-10, atol = 1e-12)

ggplot(simTyp, aes(time, Cc * 1000)) +
  geom_line() +
  geom_vline(xintercept = c(0.917, 0.917 + 2.22), linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = c(0.917 + 1.48, 0.917 + 1.48 + 1.48), linetype = "dotted", colour = "grey40") +
  labs(x = "Time after dose (h)", y = "Bedaquiline (ng/mL)",
       caption = "Dashed: depot arm (Tlag to Tlag + DUR1). Dotted: direct central arm (Tlag + Tlag,add to + DUR2).")

# Nothing enters before the tablet lag of 0.917 h. Relative floor: the ODE
# path can undershoot zero by about atol.
preLag <- simTyp$Cc[simTyp$time < 0.9]
stopifnot(all(abs(preLag) < 1e-6 * max(simTyp$Cc)))
stopifnot(max(simTyp$Cc[simTyp$time > 1]) > 0)
```

## Structural gates

Two closed-form identities check the dose split and the clarithromycin effect
independently of any published number. For a single dose the whole
bioavailable amount is eventually cleared, so `CL * AUC(0-inf) = F * Dose`
whatever FR1, the durations and the lags are; a missing dose record, a split
that does not sum to one, or a bioavailability applied to one arm only would
break it. And the clarithromycin factor on the individual clearance must be
exactly `1 - 0.37`.

```{r structural-gates}
# Terminal half-life is several months, so integrate on a log-spaced grid
# out to about 11 years and close the remainder analytically from the last
# log-linear slope.
tLong <- sort(unique(c(seq(0, 12, by = 0.01), exp(seq(log(12), log(1e5), length.out = 3000)))))
evLong <- dplyr::bind_rows(doseRows(0, 100), obsRows(tLong)) |>
  dplyr::arrange(time, dplyr::desc(evid)) |>
  dplyr::mutate(id = 1L, SEXF = 0L) |>
  addHvCov()
auc_check <- lapply(0:1, function(clr) {
  s <- rxode2::rxSolve(modTyp, dplyr::mutate(evLong, CONMED_CLARITHROMYCIN = clr),
                       returnType = "data.frame", rtol = 1e-10, atol = 1e-14,
                       maxsteps = 1e6)
  auc <- sum(diff(s$time) * (head(s$Cc, -1) + tail(s$Cc, -1)) / 2)
  n <- nrow(s)
  lz <- -diff(log(s$Cc[(n - 1):n])) / diff(s$time[(n - 1):n])
  tibble(CONMED_CLARITHROMYCIN = clr, cl = s$cl[1], fbio = s$fbio[1],
         auc_inf = auc + s$Cc[n] / lz)
}) |> dplyr::bind_rows() |>
  dplyr::mutate(ratio = cl * auc_inf / (fbio * 100))
knitr::kable(auc_check, digits = 5,
             caption = "CL x AUC(0-inf) / (F x Dose); the identity requires 1.")
stopifnot(
  all(abs(auc_check$ratio - 1) < 1e-3),
  abs(auc_check$cl[2] / auc_check$cl[1] - (1 - 0.37)) < 1e-12
)
```

# Single-dose crossover study (Kurosawa 2021 Tables 2 and 3, Figure 2)

The virtual cohort mirrors the study: healthy White adults given one 100 mg
tablet alone (A) or with steady-state clarithromycin (B), sampled to 240 h.
To make the two arms a true crossover, the same 200 subjects receive both
treatments: their random effects are drawn once in R from the model's omega
matrix and supplied as data to the typical-value model. This also makes the
cohort, and every assertion below, identical on every machine, because it does
not depend on rxode2's per-thread random-number streams.

```{r cohort}
set.seed(20210801)
omega <- ui$omega
omega <- matrix(as.numeric(omega), nrow(omega), dimnames = dimnames(omega))
drawEtas <- function(n, omega) {
  z <- matrix(stats::rnorm(n * nrow(omega)), n)
  e <- z %*% chol(omega)
  colnames(e) <- rownames(omega)
  tibble::as_tibble(e) |> dplyr::mutate(id = seq_len(n), .before = 1)
}
nSD <- 200L
etasSD <- drawEtas(nSD, omega)
sexSD <- tibble(id = seq_len(nSD), SEXF = stats::rbinom(nSD, 1, 9 / 16))

tSD <- sort(unique(c(0, 0.5, seq(1, 8, by = 0.25), 12, 24, 36, 48, 72, 96, 120, 168, 240)))
evSD <- lapply(c(A = 0L, B = 1L), function(clr) {
  one <- dplyr::bind_rows(doseRows(0, 100), obsRows(tSD)) |>
    dplyr::arrange(time, dplyr::desc(evid))
  tidyr::crossing(sexSD, one) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    dplyr::mutate(CONMED_CLARITHROMYCIN = clr) |>
    addHvCov()
})

simSD <- lapply(names(evSD), function(trt) {
  rxode2::rxSolve(modTyp, evSD[[trt]], params = etasSD, returnType = "data.frame") |>
    dplyr::mutate(treatment = trt)
}) |> dplyr::bind_rows()
```

```{r figure-2, fig.cap = "Replicates Figure 2A of Kurosawa 2021: mean (SD) bedaquiline concentration after a single 100 mg tablet, with and without clarithromycin."}
simSD |>
  dplyr::group_by(treatment, time) |>
  dplyr::summarise(mean = mean(Cc * 1000), sd = sd(Cc * 1000), .groups = "drop") |>
  dplyr::mutate(treatment = ifelse(treatment == "A", "A: bedaquiline alone",
                                   "B: with clarithromycin")) |>
  ggplot(aes(time, mean, colour = treatment)) +
  geom_line() +
  geom_errorbar(aes(ymin = pmax(mean - sd, 0), ymax = mean + sd), width = 3, alpha = 0.5) +
  labs(x = "Time after bedaquiline dose (h)", y = "Bedaquiline (ng/mL)", colour = NULL,
       caption = "Simulated, 200 virtual subjects per treatment (paired).") +
  theme(legend.position = "bottom")
```

## PKNCA

```{r pknca-sd}
concSD <- simSD |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(Cc = Cc * 1000) |>
  dplyr::select(id, time, Cc, treatment)
doseSD <- concSD |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, amt = 100)
intervalsSD <- data.frame(
  start = 0, end = c(72, 240),
  cmax = c(TRUE, FALSE), tmax = c(TRUE, FALSE), auclast = TRUE
)
ncaSD <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(concSD, Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(doseSD, amt ~ time | treatment + id),
  intervals = intervalsSD
))
resSD <- as.data.frame(ncaSD) |>
  dplyr::mutate(PPTESTCD = dplyr::case_when(
    PPTESTCD == "auclast" & end == 72 ~ "auc72",
    PPTESTCD == "auclast" & end == 240 ~ "auc240",
    TRUE ~ PPTESTCD
  ))

# Geometric means of the simulated subjects (Table 3 reports geometric means).
simGM <- resSD |>
  dplyr::filter(PPTESTCD %in% c("cmax", "auc72", "auc240")) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(PPORRES = exp(mean(log(PPORRES))), .groups = "drop") |>
  dplyr::bind_rows(
    resSD |> dplyr::filter(PPTESTCD == "tmax") |>
      dplyr::group_by(treatment, PPTESTCD) |>
      dplyr::summarise(PPORRES = median(PPORRES), .groups = "drop")
  )

# Kurosawa 2021 Table 3 geometric means (n = 16, both periods) and Table 2
# median tmax (period 1).
published <- tibble::tribble(
  ~treatment, ~cmax, ~tmax, ~auc72, ~auc240,
  "A",        1277,  5.00,  13831,  18482,
  "B",        1285,  5.00,  15458,  21065
)
cmpSD <- nlmixr2lib::ncaComparisonTable(
  simulated = simGM, reference = published, by = "treatment",
  params = c("cmax", "tmax", "auc72", "auc240"),
  units = c(cmax = "ng/mL", tmax = "h", auc72 = "ng*h/mL", auc240 = "ng*h/mL"),
  tolerance_pct = 20
)
knitr::kable(cmpSD, caption = paste(
  "Simulated vs published single-dose NCA (A: bedaquiline alone; B: with",
  "clarithromycin). Cmax and AUC are geometric means; tmax is the median.",
  "auc72 = AUC(0-72 h), auc240 = AUC(0-240 h).",
  if (!is.null(attr(cmpSD, "footnote"))) attr(cmpSD, "footnote") else ""
))

gmr <- simGM |>
  dplyr::filter(PPTESTCD %in% c("auc72", "auc240", "cmax")) |>
  tidyr::pivot_wider(names_from = treatment, values_from = PPORRES) |>
  dplyr::mutate(simulated_GMR = B / A,
                published_GMR = c(auc240 = 1.1397, auc72 = 1.1177, cmax = 1.0066)[PPTESTCD])
knitr::kable(gmr, digits = 3,
             caption = "Treatment B / A geometric mean ratio (published: Kurosawa 2021 Table 3).")

stopifnot(
  # Structural: AUC over 0-240 h sits on the published geometric mean. A
  # mis-scaled F, CL or dose would move it by far more than 20%.
  all(abs(simGM$PPORRES[simGM$PPTESTCD == "auc240"] /
            published$auc240 - 1) < 0.2),
  # Clarithromycin raises AUC(0-240 h) without changing Cmax, as published.
  gmr$simulated_GMR[gmr$PPTESTCD == "auc240"] > 1.03,
  gmr$simulated_GMR[gmr$PPTESTCD == "auc240"] < 1.25,
  abs(gmr$simulated_GMR[gmr$PPTESTCD == "cmax"] - 1) < 0.05
)
```

Simulated AUC(0-72 h) and AUC(0-240 h) are 5%-10% below the published
geometric means, and the clarithromycin ratio for AUC(0-240 h) is
`r sprintf("%.2f", gmr$simulated_GMR[gmr$PPTESTCD == "auc240"])` against the
observed 1.14. The published geometric means pool both study periods, and
Kurosawa 2021 reports carry-over of the first bedaquiline dose into period 2
(AUC(0-240 h) 8%-18% higher in period 2 than in period 1, Table 2), which a
single-dose simulation does not include. Simulated Cmax is about 16% lower than
observed and the simulated median tmax is earlier (about 4 h vs 5 h; the
starred row); Kurosawa 2021 notes that the updated model had its largest
conditional weighted residuals in the absorption phase (1-2 h after dosing).
None of these differences is tuned away.

# Steady-state regimens (Kurosawa 2021 Table 5 and Figure 3)

Kurosawa 2021 simulated 48 weeks of bedaquiline in non-Black adults (1:1
male:female) with a disease status similar to MDR-TB. The reference regimen is
the MDR-TB regimen without clarithromycin (400 mg once daily for 2 weeks, then
200 mg three times a week); regimens A-D add clarithromycin:

* A: 400 mg once daily for 2 weeks, then 200 mg twice a week
* B: 400 mg once daily for 2 weeks, then 100 mg three times a week
* C: 400 mg once daily for 2 weeks, then 100 mg twice a week
* D: 400 mg once daily for 2 weeks, then 100 mg five times a week

The simulation below uses the same 200 subjects for every regimen, tablet
formulation, `DIS_TB_MDR = 1` (no healthy-volunteer clearance increase) and the
**reference-study bioavailability, F = 1** (`STUDY_BDQ_C208_C209 = 1`). The
choice of F is discussed under *Assumptions and deviations*: with the 2.03 of
the "other studies" group every simulated exposure is almost exactly 2.03
times Table 5.

```{r cohort-ss}
set.seed(20210802)
nSS <- 200L
etasSS <- drawEtas(nSS, omega)
sexSS <- tibble(id = seq_len(nSS), SEXF = rep(0:1, length.out = nSS))

# Loading: 400 mg daily on days 0-13. Maintenance from week 3: dose days
# within each week (Monday = 0).
regimens <- tibble(
  regimen = c("MDR-TB (ref.)", "A", "B", "C", "D"),
  maint_amt = c(200, 200, 100, 100, 100),
  days = list(c(0, 2, 4), c(0, 3), c(0, 2, 4), c(0, 3), 0:4),
  clr = c(0L, 1L, 1L, 1L, 1L)
)
tObsSS <- sort(unique(c(
  (1:48) * 168,                    # weekly trough, end of each week
  seq(13 * 24, 14 * 24, by = 0.25), # day 14 (week 2)
  seq(23 * 168, 24 * 168, by = 1), # week 24
  seq(47 * 168, 48 * 168, by = 1)  # week 48
)))
evSS <- lapply(seq_len(nrow(regimens)), function(i) {
  maint <- unlist(lapply(2:47, function(w) (w * 7 + regimens$days[[i]]) * 24))
  one <- dplyr::bind_rows(
    doseRows(c((0:13) * 24, maint), c(rep(400, 14), rep(regimens$maint_amt[i], length(maint)))),
    obsRows(tObsSS)
  ) |> dplyr::arrange(time, dplyr::desc(evid))
  tidyr::crossing(sexSS, one) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    dplyr::mutate(RACE_BLACK = 0L, DIS_TB_MDR = 1L, STUDY_BDQ_C208_C209 = 1L,
                  STUDY_BDQ_CDE102_C104 = 0L, FORM_TABLET = 1L,
                  CONMED_CLARITHROMYCIN = regimens$clr[i])
})
names(evSS) <- regimens$regimen
simSS <- lapply(names(evSS), function(r) {
  rxode2::rxSolve(modTyp, evSS[[r]], params = etasSS, returnType = "data.frame") |>
    dplyr::mutate(regimen = r)
}) |> dplyr::bind_rows()
```

```{r figure-3, fig.cap = "Replicates Figure 3 of Kurosawa 2021: simulated mean bedaquiline trough concentration by week."}
simSS |>
  dplyr::filter(time %in% ((1:48) * 168)) |>
  dplyr::group_by(regimen, week = time / 168) |>
  dplyr::summarise(ctrough = mean(Cc * 1000), .groups = "drop") |>
  ggplot(aes(week, ctrough, colour = regimen)) +
  geom_line() +
  labs(x = "Week", y = "Mean Ctrough (ng/mL)", colour = NULL,
       caption = "MDR-TB (ref.) is without clarithromycin; A-D are with clarithromycin.")
```

## PKNCA against Table 5

```{r pknca-ss}
concSS <- simSS |>
  dplyr::filter(!is.na(Cc), regimen %in% c("MDR-TB (ref.)", "A", "D")) |>
  dplyr::mutate(Cc = Cc * 1000) |>
  dplyr::select(id, time, Cc, regimen)
doseSS <- dplyr::bind_rows(lapply(c("MDR-TB (ref.)", "A", "D"), function(r) {
  evSS[[r]] |> dplyr::filter(evid == 1L, cmt == "depot") |>
    dplyr::select(id, time, amt) |> dplyr::mutate(regimen = r)
}))
# Week 2 = day 14 (AUC0-24); weeks 24 and 48 = AUC over the whole week.
intervalsSS <- data.frame(
  start = c(13 * 24, 23 * 168, 47 * 168),
  end = c(14 * 24, 24 * 168, 48 * 168),
  cmax = TRUE, auclast = TRUE
)
ncaSS <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(concSS, Cc ~ time | regimen + id),
  PKNCA::PKNCAdose(doseSS, amt ~ time | regimen + id),
  intervals = intervalsSS
))
simT5 <- as.data.frame(ncaSS) |>
  dplyr::mutate(week = c(`312` = "2", `3864` = "24", `7896` = "48")[as.character(start)]) |>
  dplyr::group_by(regimen, week, PPTESTCD) |>
  dplyr::summarise(PPORRES = mean(PPORRES), .groups = "drop") |>
  # Ctrough is the single concentration at the end of each interval (the
  # next dose starts at that time, after its lag), read directly from the
  # simulation rather than through PKNCA.
  dplyr::bind_rows(
    concSS |>
      dplyr::filter(time %in% c(14 * 24, 24 * 168, 48 * 168)) |>
      dplyr::mutate(week = c(`336` = "2", `4032` = "24", `8064` = "48")[as.character(time)]) |>
      dplyr::group_by(regimen, week) |>
      dplyr::summarise(PPORRES = mean(Cc), .groups = "drop") |>
      dplyr::mutate(PPTESTCD = "ctrough")
  )

# Kurosawa 2021 Table 5, simulated means (n = 1000 per regimen).
table5 <- tibble::tribble(
  ~regimen,        ~week, ~ctrough, ~cmax, ~auclast,
  "MDR-TB (ref.)", "2",   1125,     3274,  42652,
  "MDR-TB (ref.)", "24",  850,      2013,  180526,
  "MDR-TB (ref.)", "48",  1069,     2233,  217615,
  "A",             "2",   1262,     3372,  45900,
  "A",             "24",  794,      1919,  161430,
  "A",             "48",  1009,     2137,  197865,
  "D",             "2",   1258,     3358,  45780,
  "D",             "24",  944,      1596,  191203,
  "D",             "48",  1230,     1884,  239657
)
cmpT5 <- nlmixr2lib::ncaComparisonTable(
  simulated = simT5, reference = table5, by = c("regimen", "week"),
  units = c(ctrough = "ng/mL", cmax = "ng/mL", auclast = "ng*h/mL"),
  tolerance_pct = 20
)
knitr::kable(cmpT5, caption = paste(
  "Simulated vs Kurosawa 2021 Table 5 mean exposures. auclast is AUC(0-24 h)",
  "at week 2 and AUC(0-168 h) at weeks 24 and 48.",
  if (!is.null(attr(cmpT5, "footnote"))) attr(cmpT5, "footnote") else ""
))

pctT5 <- simT5 |>
  dplyr::inner_join(
    tidyr::pivot_longer(table5, c(ctrough, cmax, auclast),
                        names_to = "PPTESTCD", values_to = "ref"),
    by = c("regimen", "week", "PPTESTCD")
  ) |>
  dplyr::mutate(pct = 100 * (PPORRES / ref - 1))
stopifnot(
  nrow(pctT5) == 27L,
  !anyNA(pctT5$pct),
  # Structural: F, CL, the regimen and the clarithromycin effect together
  # put the 27 means on Table 5. With F = 2.03 the median would be +100%.
  abs(median(pctT5$pct)) < 8,
  # Envelope, robust to which subjects land in the tails.
  quantile(abs(pctT5$pct), 0.9) < 15
)
```

The 27 simulated means match Table 5 closely (median difference
`r sprintf("%.1f", median(pctT5$pct))`%, largest
`r sprintf("%.1f", max(abs(pctT5$pct)))`% in absolute value), including the
week-24 Cmax deficit of regimen D that Kurosawa 2021 singles out.

## The between-subject variability scale

Supplemental Table S1 lists the between-subject variability as bare numbers
(50.4 for CL/F, 39.1 for Vc/F, 113 for FR1, 39.6 for F) under a "BSV" header,
and McLeay 2014 calls them "%CV". That leaves two readings: the number is the
SD of eta (omega = 0.504), or a back-transformed CV (omega^2 = log(1 + 0.504^2)).
FR1 carries its eta on the logit scale, where a back-transformed CV has no
meaning, which already favours the first. Table 5 supports it: the SD / mean
of the simulated trough and AUC depends on the clearance and bioavailability
omegas, and the eta-SD reading reproduces the published spread more closely
than the log(1 + CV^2) reading, which is too narrow. The same standard-normal
draws are used for both readings, so the comparison is paired. (Cmax also
depends on the large logit-scale FR1 variance and does not separate the two
readings at 200 subjects; in a 1,000-subject check made by the maintainers
while building the model, the eta-SD reading was the closer one for all nine
Table 5 reference-regimen cells.)

```{r omega-scale}
cv <- sqrt(diag(omega))
rho <- omega["etalcl", "etalvc"] / (cv["etalcl"] * cv["etalvc"])
omegaLog <- diag(log(1 + cv^2))
dimnames(omegaLog) <- dimnames(omega)
omegaLog["etalcl", "etalvc"] <- omegaLog["etalvc", "etalcl"] <-
  rho * sqrt(omegaLog["etalcl", "etalcl"] * omegaLog["etalvc", "etalvc"])
# Rescale the SAME eta draws to the alternative omega.
zSS <- as.matrix(etasSS[, rownames(omega)]) %*% solve(chol(omega))
etasLog <- tibble::as_tibble(zSS %*% chol(omegaLog)) |>
  dplyr::mutate(id = etasSS$id, .before = 1)

simLog <- rxode2::rxSolve(modTyp, evSS[["MDR-TB (ref.)"]], params = etasLog,
                          returnType = "data.frame")
spread <- function(s, reading) {
  wk <- function(lo, hi, lab) s |> dplyr::filter(time >= lo, time <= hi) |>
    dplyr::group_by(id) |>
    dplyr::summarise(ctrough = Cc[which.max(time)], cmax = max(Cc),
                     auc = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2)) |>
    dplyr::summarise(dplyr::across(c(ctrough, cmax, auc), ~ sd(.x) / mean(.x))) |>
    dplyr::mutate(week = lab)
  dplyr::bind_rows(wk(312, 336, "2"), wk(3864, 4032, "24"), wk(7896, 8064, "48")) |>
    dplyr::mutate(reading = reading)
}
spreadTab <- dplyr::bind_rows(
  spread(dplyr::filter(simSS, regimen == "MDR-TB (ref.)"), "omega = BSV/100 (packaged)"),
  spread(simLog, "omega^2 = log(1 + CV^2)"),
  tibble(ctrough = c(523 / 1125, 489 / 850, 667 / 1069),
         cmax = c(1657 / 3274, 1059 / 2013, 1212 / 2233),
         auc = c(19684 / 42652, 96964 / 180526, 126252 / 217615),
         week = c("2", "24", "48"), reading = "Kurosawa 2021 Table 5")
)
knitr::kable(spreadTab, digits = 3,
             caption = "SD / mean of the MDR-TB reference-regimen exposures.")

gap <- spreadTab |>
  tidyr::pivot_longer(c(ctrough, cmax, auc)) |>
  tidyr::pivot_wider(names_from = reading, values_from = value) |>
  dplyr::mutate(
    sdReading = abs(`omega = BSV/100 (packaged)` - `Kurosawa 2021 Table 5`),
    logReading = abs(`omega^2 = log(1 + CV^2)` - `Kurosawa 2021 Table 5`)
  )
# Trough and AUC: the packaged reading is the closer one on average (measured
# 0.027 vs 0.049 mean absolute gap in SD / mean).
gapExposure <- dplyr::filter(gap, name %in% c("ctrough", "auc"))
stopifnot(mean(gapExposure$sdReading) < mean(gapExposure$logReading))
```

# Assumptions and deviations

* **Bioavailability in the Table 5 simulations.** Supplemental Table S1
  footnote a lists "Other studies on F" (2.03) among the covariates used in
  the simulations. Simulated with F = 2.03, every Table 5 mean is reproduced
  almost exactly 2.03-fold too high (in a 1,000-subject check by the
  maintainers, week-24 AUC(0-168 h) of the reference regimen was about
  353,000-362,000 vs 180,526 ng*h/mL), while F = 1,
  the McLeay 2014 reference for the MDR-TB phase IIb studies, reproduces all
  27 means to within 10% (table above). Because the model is linear, no other parameter can produce a
  uniform 2-fold shift across trough, peak and AUC, so the article's
  simulations are taken to have used F = 1, consistent with its stated
  assumption of an MDR-TB-like disease status. The packaged model keeps the
  full study covariate; use `STUDY_BDQ_C208_C209 = 1` to reproduce the
  article's simulations and both study indicators = 0 for a new healthy
  volunteer study (as for the single-dose crossover data).
* **Clearance with clarithromycin.** Kurosawa 2021 Results report CL/F of
  1.81 L/h with clarithromycin vs 2.78 L/h alone. `2.78 x (1 - 0.37)` is
  1.75 L/h; 1.81 / 2.78 corresponds to -35%. The tabulated -0.37 (RSE 11%) and
  its 95% CI (-45% to -29%) are consistent with each other and are used.
* **Healthy-volunteer clearance.** The +37.5% CL/F for healthy volunteers and
  drug-sensitive TB patients (McLeay 2014) is encoded on `1 - DIS_TB_MDR`. The
  article's simulations did not apply it (Table S1 footnote a); the single-dose
  crossover simulation here does, because those subjects were healthy
  volunteers (Table S1 footnote b lists it among the covariates used for their
  post hoc estimates).
* **Lag of the direct central input.** McLeay 2014 Fig. 1 labels the lag of the
  second, direct-to-central arm "Tlag + Tlag,add", so the arm starts 1.48 h
  after the formulation lag, not after the end of the depot input.
* **Duplicated 1.48 h.** Tlag,add and DUR2 are both 1.48 h with identical RSE
  (3.2%) in McLeay 2014 Table 3, suggesting they shared one parameter in the
  original fit. They are encoded as two fixed parameters with the same value,
  which is equivalent for simulation.
* **Residual error.** McLeay 2014 fitted log-transformed concentrations with an
  additive error (log-transform-both-sides), encoded as `lnorm()`; the value
  printed as "CV%" is used as the log-scale SD, consistent with the
  between-subject variability reading above. Residual error is not used in any
  simulation here.
* **Virtual cohorts.** Both cohorts are 200 subjects, with sex drawn to match
  Table 1 (single-dose crossover; 9 of 16 women) or 1:1 (Table 5, as the
  article specifies); no Black subjects, as in both the study and the
  article's simulations.
* **Errata.** No correction notice for Kurosawa 2021 or McLeay 2014 was found
  as of 2026-09-29.
