---
title: "MG1113 anti-TFPI antibody two-target TMDD across species (Kwak 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MG1113 anti-TFPI antibody two-target TMDD across species (Kwak 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Kwak 2025 refines a target-mediated drug disposition (TMDD) model for **MG1113**,
a humanized IgG4 antibody against the Kunitz-type protease inhibitor 2 (K2) domain
of tissue factor pathway inhibitor (TFPI), in development for hemophilia. The
refinement over the earlier Kwak 2021 model is threefold: both TFPI pools are
represented explicitly (soluble TFPI-alpha *and* the far more abundant
membrane-bound TFPI), the empirical feedback compartment used previously to
generate the sTFPI-alpha rebound is replaced by a mechanistic one, and a transit
compartment is added for delayed subcutaneous absorption.

The paper contributes **three model files** -- one fit and two allometric
projections:

| Model | Species | Provenance |
|---|---|---|
| `Kwak_2025_mg1113_monkey` | cynomolgus monkey | Fitted to monkey data by the Cluster Gauss-Newton Method (Table 1, rank 1) |
| `Kwak_2025_mg1113_rabbit` | rabbit | Allometric projection of the monkey fit; externally validated (Table 3, Figure 3) |
| `Kwak_2025_mg1113_human`  | human  | Allometric projection of the monkey fit; **not validated** (Table 3, Figure 4) |

* Article: <https://doi.org/10.3389/fphar.2025.1745702>
* Supplement (model equations, Table S1-S4): <https://www.frontiersin.org/articles/10.3389/fphar.2025.1745702/full#supplementary-material>

```{r load-models}
models <- list(
  monkey = rxode2::rxode(readModelDb("Kwak_2025_mg1113_monkey")),
  rabbit = rxode2::rxode(readModelDb("Kwak_2025_mg1113_rabbit")),
  human  = rxode2::rxode(readModelDb("Kwak_2025_mg1113_human"))
)
cat(models$monkey$description)
```

# Population

The structural model was fitted to previously published plasma MG1113 and
sTFPI-alpha profiles from **cynomolgus monkeys** (Kwak 2025 section 2.1,
reanalysing Kwak 2021). Monkeys received a single intravenous or subcutaneous
dose of MG1113 at 2.5, 5.0 or 10.0 mg/kg (17.2, 34.4 or 68.8 nmol/kg), with
n = 1-3 per dose/route group. Individual body weights were unavailable, so a
representative 3.5 kg was used throughout and doses were converted to 60.2,
120.4 and 240.8 nmol per animal. Both analytes were measured by ELISAs detecting
the **free** forms, and concentrations were converted to molar units using
molecular weights of 145.36 kDa (MG1113) and 35 kDa (sTFPI-alpha).

The **rabbit** cohort (male, 2-3 months, 2.5 kg representative weight) received
single i.v. doses of 2.75, 17.2 or 34.4 nmol/kg or single s.c. doses of 17.2,
34.4, 68.8 or 137.6 nmol/kg, n = 2-10 per group. These data were used only to
*evaluate* the allometric projection, never to fit it.

No **human** MG1113 data were available to the authors, so the human file is a
pure forward prediction under the phase I regimens of NCT03855696 (s.c. 0.5, 1.7
and 3.3 mg/kg; i.v. 3.3 mg/kg).

The same information is available programmatically via each model's `population`
metadata, e.g. `readModelDb("Kwak_2025_mg1113_monkey")()$population`.

```{r population}
tibble::tibble(
  Model   = names(models),
  Species = vapply(models, function(m) m$population$species, character(1)),
  Dosing  = vapply(models, function(m) m$population$dose_range, character(1))
) |>
  knitr::kable(caption = "Study populations behind each of the three model files.")
```

# Model structure

All three files share one structure (Kwak 2025 Figure 1; equations in
Supplementary Material section 4). Writing `C2` for the free MG1113
concentration in the central compartment:

* **Absorption (s.c. only).** The depot empties at `k01` into a transit
  compartment, at `k02` directly into plasma, and is lost at `kloss`. The
  transit compartment feeds plasma at `k12`.
* **Disposition.** Two-compartment distribution with inter-compartmental
  clearance `CLD` between volumes `V2` and `V3`; first-order elimination of free
  drug at `kel,MG1113`.
* **Two-target TMDD.** MG1113 binds free sTFPI-alpha and free mTFPI with a
  common association constant `kon`; `koff = KD * kon` with `KD` fixed from the
  in-vitro measurement. Each target turns over with zero-order synthesis and
  first-order degradation, and each drug-target complex is eliminated by its own
  first-order constant (`kel,CS` and `kel,CM`).

Two structural notes on the encoding:

1. The paper writes its central and peripheral ODEs on **concentration**; the
   model files carry those two states as **amounts** instead, so that a dose can
   land on `central` (i.v.) or `depot` (s.c.) through an ordinary rxode2 event
   table. Each ODE is the paper's equation multiplied through by its compartment
   volume, so the two forms are identical. The four TFPI-species states
   (`stfpi`, `mtfpi`, `astfpi`, `amtfpi`) remain concentrations in nM, exactly as
   the paper writes them.
2. Neither synthesis rate is a free parameter: both are pinned by the
   steady-state balance at the reported baseline, `ksyn = kdeg * baseline`
   (Supplementary Material section 4, "Related equations"). Likewise `koff` is
   pinned by `KD`.

# Source trace

Per-parameter provenance is recorded as an in-file comment beside each `ini()`
entry. Collected here for review:

| Equation / parameter | Monkey value | Source location |
|---|---|---|
| `KD` | 0.04665 nM (fixed) | Table 1, "Fixed parameter" block (from Kwak 2020) |
| `kon` | 28.50 1/(nM*day) | Table 1, rank 1 |
| `koff = KD * kon` | derived | Suppl. Material section 4, "Related equations" |
| `lq` (CLD) | 0.1454 L/day | Table 1, rank 1 |
| `lvc` (V2) | 0.1112 L | Table 1, rank 1 |
| `lvp` (V3) | 0.1346 L | Table 1, rank 1 |
| `k_depot_transit1` (k01) | 20.32 /day | Table 1, rank 1 |
| `k_depot_central` (k02) | 8.859e-8 /day | Table 1, rank 1 |
| `k_transit1_central` (k12) | 0.4033 /day | Table 1, rank 1 |
| `kloss` | 3.149e-7 /day | Table 1, rank 1 |
| `lkel` (kel,MG1113) | 0.4543 /day | Table 1, rank 1 |
| `kdegs` (kdeg,s) | 75.50 /day | Table 1, rank 1 |
| `kdegm` (kdeg,m) | 1.150 /day | Table 1, rank 1 |
| `kints` (kel,CS) | 0.3094 /day | Table 1, rank 1 |
| `kintm` (kel,CM) | 0.006851 /day | Table 1, rank 1 |
| `stfpi_b` | 0.9456 nM | Table 1, rank 1 |
| `mtfpi_b` | 12.04 nM | Table 1, rank 1 |
| `ksyn = kdeg * baseline` | derived | Suppl. Material section 4, "Related equations" |
| all `d/dt(...)` equations | n/a | Suppl. Material section 4, "Model equations" |
| rabbit / human parameter columns | see Table 3 | Table 3; allometry per section 2.6 |

# Simulation helpers

Every model here is a **typical-value deterministic simulator**: the CGNM fit
targeted the mean observed profiles, so the paper reports no inter-individual
variability and no residual-error model, and none is encoded. There is therefore
no cohort to draw and no VPC to build -- one subject per arm reproduces the
published curves exactly, and every assertion below is on a deterministic
quantity, so tight bounds are appropriate.

```{r helpers}
MW_MG1113 <- 145360                      # g/mol, Kwak 2025 section 2.1
nM_to_ugmL <- MW_MG1113 * 1e-9 * 1e3     # 1 nM = 0.14536 ug/mL

# A time grid dense enough to resolve the steep early distribution phase (an
# under-resolved grid understates AUC by several percent; see pattern 11 of
# the known-vignette-failure-patterns reference).
obs_grid <- function(tmax) {
  sort(unique(c(
    seq(0, 1, by = 0.002),
    seq(1, 10, by = 0.01),
    seq(10, tmax, by = 0.05)
  )))
}

# Build an explicit event table. Observation rows carry cmt = "central" -- an
# ODE STATE, never the observable name "Cc" (that would auto-inject a cmt()
# slot and renumber every compartment).
make_events <- function(amt, route, tmax, ii = NULL, addl = 0L, id = 1L) {
  dose_cmt <- if (route == "iv") "central" else "depot"
  dose_times <- if (is.null(ii)) 0 else seq(0, by = ii, length.out = addl + 1L)
  dplyr::bind_rows(
    tibble::tibble(id = id, time = dose_times, amt = amt, evid = 1L, cmt = dose_cmt),
    tibble::tibble(id = id, time = obs_grid(tmax), amt = NA_real_, evid = 0L, cmt = "central")
  ) |>
    dplyr::arrange(time, dplyr::desc(evid))
}

solve_arm <- function(model, events, keep = character()) {
  # useLinCmt = FALSE: rxode2's automatic ODE->linCmt conversion corrupts the
  # observable mapping for multi-output multi-state models.
  out <- rxode2::rxSolve(model, events = events, keep = keep, useLinCmt = FALSE) |>
    as.data.frame()
  if (is.null(out$id)) out$id <- 1L
  out
}

# Trapezoidal AUC on a solved arm.
auc_of <- function(df, col, from = 0, to = Inf) {
  d <- df[df$time >= from & df$time <= to, ]
  d <- d[order(d$time), ]
  sum(diff(d$time) * (utils::head(d[[col]], -1) + utils::tail(d[[col]], -1)) / 2)
}
```

# Structural check 1: the undosed system sits exactly at baseline

`ksyn = kdeg * baseline` should hold each target perfectly flat in the absence of
drug. This is a deterministic identity, so the tolerance is machine precision,
not a tuned bound -- it goes red on any mis-transcribed baseline, degradation
rate or synthesis expression.

```{r ss-check}
ss <- lapply(names(models), function(sp) {
  m  <- models[[sp]]
  ev <- tibble::tibble(id = 1L, time = seq(0, 60, by = 0.25),
                       amt = NA_real_, evid = 0L, cmt = "central")
  s  <- solve_arm(m, ev)
  b  <- m$theta
  tibble::tibble(
    Species          = sp,
    `sTFPI-a base (nM)` = unname(b[["stfpi_b"]]),
    `max |dev|`      = max(abs(s$sTFPI - unname(b[["stfpi_b"]]))),
    `mTFPI base (nM)` = unname(b[["mtfpi_b"]]),
    `max |dev| mTFPI` = max(abs(s$mtfpi - unname(b[["mtfpi_b"]])))
  )
}) |> dplyr::bind_rows()

knitr::kable(ss, digits = 12,
             caption = "Undosed steady state: both targets hold their baseline exactly.")

stopifnot(all(ss$`max |dev|` < 1e-6), all(ss$`max |dev| mTFPI` < 1e-6))
```

# Structural check 2: the allometric scaling reproduces Table 3

Table 3 is the monkey rank 1 column scaled to 2.5 kg (rabbit) and 70 kg (human)
with exponents 0.75 for `CLD`, -0.25 for every rate constant and 1.0 for the
volumes (Kwak 2025 section 2.6). Recomputing that scaling from the monkey file
and comparing against the values actually encoded in the rabbit and human files
checks the whole Table 3 transcription in one step.

```{r allometry}
BW <- c(monkey = 3.5, rabbit = 2.5, human = 70.0)   # Table 3
expo <- c(lq = 0.75, lvc = 1.0, lvp = 1.0,
          k_depot_transit1 = -0.25, k_depot_central = -0.25,
          k_transit1_central = -0.25, lkel = -0.25,
          kdegs = -0.25, kdegm = -0.25, kints = -0.25, kintm = -0.25,
          kloss = -0.25)
# Parameters held on the log scale in ini() must be scaled on the linear scale.
log_scale <- c("lq", "lvc", "lvp", "lkel")

theta_lin <- function(m) {
  th <- m$theta
  for (p in log_scale) th[[p]] <- exp(th[[p]])
  th
}
mk <- theta_lin(models$monkey)

allo <- lapply(c("rabbit", "human"), function(sp) {
  tgt <- theta_lin(models[[sp]])
  ratio <- BW[[sp]] / BW[["monkey"]]
  tibble::tibble(
    Species   = sp,
    Parameter = names(expo),
    Expected  = unname(mk[names(expo)]) * ratio^unname(expo),
    Encoded   = unname(tgt[names(expo)])
  )
}) |>
  dplyr::bind_rows() |>
  dplyr::mutate(`% diff` = 100 * (Encoded - Expected) / Expected)

allo |>
  dplyr::mutate(dplyr::across(c(Expected, Encoded), ~ signif(.x, 4)),
                `% diff` = round(`% diff`, 2)) |>
  knitr::kable(caption = paste(
    "Table 3 check: each rabbit / human value recomputed from the monkey fit",
    "and the paper's allometric exponents, vs the value encoded in the file."))

# The paper rounds Table 3 to 4 significant figures, so exact agreement is not
# expected; anything beyond ~0.5% would mean a mis-transcribed column.
stopifnot(max(abs(allo$`% diff`)) < 0.5)
```

The two target baselines are deliberately **excluded** from that check: they are
not allometrically scaled. `stfpi_b` is a per-species literature value (0.9456 nM
fitted in monkeys, 1.114 nM in rabbits from Kwak 2020, 2.3 nM in humans from
Kwak 2021) and `mtfpi_b` follows from the monkey mTFPI:sTFPI-alpha ratio of 12.7
(Supplementary Material item 7).

```{r baselines}
tibble::tibble(
  Species = names(models),
  `sTFPI-a base (nM)` = vapply(models, function(m) unname(m$theta[["stfpi_b"]]), numeric(1)),
  `mTFPI base (nM)`   = vapply(models, function(m) unname(m$theta[["mtfpi_b"]]), numeric(1))
) |>
  dplyr::mutate(Ratio = round(`mTFPI base (nM)` / `sTFPI-a base (nM)`, 2)) |>
  knitr::kable(caption = "Target baselines are per-species literature values, not allometric projections.")
```

# Replicating Figure 2: monkey MG1113 and sTFPI-alpha

```{r figure-2, fig.height = 6}
monkey_doses <- c("17.2 nmol/kg" = 60.2, "34.4 nmol/kg" = 120.4, "68.8 nmol/kg" = 240.8)

monkey_sim <- do.call(dplyr::bind_rows, lapply(c("iv", "sc"), function(rt) {
  do.call(dplyr::bind_rows, lapply(names(monkey_doses), function(dl) {
    solve_arm(models$monkey, make_events(monkey_doses[[dl]], rt, tmax = 28)) |>
      dplyr::mutate(route = toupper(rt), dose = dl)
  }))
}))

monkey_sim |>
  dplyr::filter(time > 0) |>
  tidyr::pivot_longer(c(Cc, sTFPI), names_to = "analyte", values_to = "conc") |>
  dplyr::mutate(analyte = dplyr::recode(analyte, Cc = "MG1113", sTFPI = "sTFPI-alpha")) |>
  ggplot(aes(time, conc, colour = dose)) +
  geom_line(linewidth = 0.7) +
  facet_grid(analyte ~ route, scales = "free_y") +
  scale_y_log10() +
  labs(x = "Time (day)", y = "Plasma concentration (nM)", colour = "Dose",
       title = "Figure 2 - monkey MG1113 and sTFPI-alpha profiles",
       caption = "Replicates Figure 2 of Kwak 2025 (rank 1 parameter set).") +
  theme_bw()
```

The mechanistic point of the refinement is the **sTFPI-alpha rebound**: Aston
2014/2017 show a rebound arises when the complex internalisation rate is slower
than both the free-drug elimination rate and the free-target degradation rate.
The monkey fit satisfies exactly that ordering, and the simulated profile
overshoots baseline as a result.

```{r rebound}
b   <- models$monkey$theta
ord <- c(`kel,CS (complex)` = unname(b[["kints"]]),
         `kel,MG1113 (free drug)` = exp(unname(b[["lkel"]])),
         `kdeg,s (free target)` = unname(b[["kdegs"]]))

hi  <- monkey_sim |> dplyr::filter(route == "IV", dose == "68.8 nmol/kg")
nadir_t <- hi$time[which.min(hi$sTFPI)]
peak    <- max(hi$sTFPI[hi$time > nadir_t])
base    <- unname(b[["stfpi_b"]])

knitr::kable(
  tibble::tibble(Quantity = names(ord), `Value (1/day)` = signif(unname(ord), 4)),
  caption = "Aston rebound condition: kel,CS must be below BOTH of the other two rates.")

cat(sprintf(
  "IV 68.8 nmol/kg: baseline %.4f nM, nadir %.4g nM at day %.2f, post-nadir peak %.4f nM (%.1f%% of baseline).\n",
  base, min(hi$sTFPI), nadir_t, peak, 100 * peak / base))

stopifnot(
  # The rebound condition the paper argues from (Discussion, section 4).
  ord[["kel,CS (complex)"]] < ord[["kel,MG1113 (free drug)"]],
  ord[["kel,CS (complex)"]] < ord[["kdeg,s (free target)"]],
  # ... and the rebound it predicts: sTFPI-alpha overshoots its own baseline.
  peak > base
)
```

# Replicating Figure 3: the rabbit projection

```{r figure-3}
rabbit_arms <- tibble::tribble(
  ~route, ~label,           ~nmol_per_kg,
  "iv",   "IV 2.75",        2.75,
  "iv",   "IV 17.2",        17.2,
  "iv",   "IV 34.4",        34.4,
  "sc",   "SC 17.2",        17.2,
  "sc",   "SC 34.4",        34.4,
  "sc",   "SC 68.8",        68.8,
  "sc",   "SC 137.6",       137.6
) |>
  dplyr::mutate(amt = nmol_per_kg * 2.5)   # 2.5 kg representative rabbit

rabbit_sim <- do.call(dplyr::bind_rows, lapply(seq_len(nrow(rabbit_arms)), function(i) {
  solve_arm(models$rabbit, make_events(rabbit_arms$amt[i], rabbit_arms$route[i], tmax = 21)) |>
    dplyr::mutate(route = toupper(rabbit_arms$route[i]), arm = rabbit_arms$label[i])
}))

rabbit_sim |>
  dplyr::filter(time > 0) |>
  tidyr::pivot_longer(c(Cc, sTFPI), names_to = "analyte", values_to = "conc") |>
  dplyr::mutate(analyte = dplyr::recode(analyte, Cc = "MG1113", sTFPI = "sTFPI-alpha")) |>
  ggplot(aes(time, conc, colour = arm)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~ analyte, scales = "free_y") +
  scale_y_log10() +
  labs(x = "Time (day)", y = "Plasma concentration (nM)", colour = "Arm",
       title = "Figure 3 - rabbit projection under the observed dosing arms",
       caption = "Replicates Figure 3 of Kwak 2025. Parameters are allometric predictions, not a rabbit fit.") +
  theme_bw()
```

The paper compares these predictions with observed rabbit data and reports
absolute average fold errors of 1.5-1.9 in the high-dose groups but 2.8-4.2 in
the low-dose groups, where the model over-predicts and misses a rapid decline
(section 3.2). The rabbit observations themselves are not published as a table,
so that comparison cannot be reproduced numerically here; it is recorded as a
known limitation of the rabbit file rather than reproduced.

# Replicating Figure 4: the human projection

```{r figure-4, fig.height = 6}
human_doses_mgkg <- c("0.5 mg/kg" = 0.5, "1.7 mg/kg" = 1.7, "3.3 mg/kg" = 3.3)
# Kwak 2025 section 3.3: 0.5, 1.7, 3.3 mg/kg = 3.44, 11.7, 22.7 nmol/kg
human_doses_nmol <- c("0.5 mg/kg" = 3.44, "1.7 mg/kg" = 11.7, "3.3 mg/kg" = 22.7) * 70

human_single <- do.call(dplyr::bind_rows, lapply(c("iv", "sc"), function(rt) {
  do.call(dplyr::bind_rows, lapply(names(human_doses_nmol), function(dl) {
    solve_arm(models$human, make_events(human_doses_nmol[[dl]], rt, tmax = 30)) |>
      dplyr::mutate(route = toupper(rt), dose = dl)
  }))
}))

human_q7d <- do.call(dplyr::bind_rows, lapply(c("iv", "sc"), function(rt) {
  do.call(dplyr::bind_rows, lapply(names(human_doses_nmol), function(dl) {
    solve_arm(models$human,
              make_events(human_doses_nmol[[dl]], rt, tmax = 56, ii = 7, addl = 7L)) |>
      dplyr::mutate(route = toupper(rt), dose = dl)
  }))
}))

stfpi_base_human <- unname(models$human$theta[["stfpi_b"]])

dplyr::bind_rows(
  dplyr::mutate(human_single, panel = "Single dose"),
  dplyr::mutate(human_q7d,    panel = "Q7d x 8")
) |>
  dplyr::filter(time > 0) |>
  ggplot(aes(time, sTFPI, colour = dose)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 0.25 * stfpi_base_human, linetype = "dotted") +
  facet_grid(panel ~ route, scales = "free_x") +
  labs(x = "Time (day)", y = "Plasma sTFPI-alpha (nM)", colour = "Dose",
       title = "Figure 4 - predicted human sTFPI-alpha suppression",
       caption = paste("Replicates Figure 4B/4D of Kwak 2025. Dotted line = 25% of the",
                       "2.3 nM baseline, the efficacy-associated target.")) +
  theme_bw()
```

# PKNCA validation against Supplementary Table S3

Supplementary Table S3 tabulates the model's own predicted human
AUC<sub>0-30days</sub> for all six single-dose arms, in day*ug/mL. Because the
integration window is stated exactly (0 to 30 days), this is an **exact**
validation target rather than an approximate one: a correct re-implementation
must reproduce all six numbers. Any error in an ODE term, a parameter value, the
molar dose conversion or the nM-to-ug/mL conversion moves them.

```{r pknca}
sim_nca <- human_single |>
  dplyr::mutate(arm = paste(route, dose)) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

# Guarantee a time = 0 row per arm so PKNCA can anchor AUC0-30. IDs are shared
# across arms here (one deterministic subject per arm), so the arm label -- not
# id -- is the grouping that matters; give each arm its own id to be safe.
sim_nca <- sim_nca |>
  dplyr::mutate(id = as.integer(factor(arm)))

sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, arm, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

dose_df <- human_single |>
  dplyr::mutate(arm = paste(route, dose)) |>
  dplyr::distinct(arm, dose, route) |>
  dplyr::mutate(id = as.integer(factor(arm)), time = 0,
                amt = unname(human_doses_nmol[dose]))

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)

intervals <- data.frame(
  start = 0, end = 30,
  auclast = TRUE, cmax = TRUE, tmax = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r nca-compare}
# Supplementary Table S3, single-administration block (day*ug/mL), converted to
# the model's native day*nM so the comparison stays in the units the model emits.
published <- tibble::tribble(
  ~arm,          ~auclast_ugmL,
  "IV 0.5 mg/kg",   4.018,
  "IV 1.7 mg/kg",  40.16,
  "IV 3.3 mg/kg", 119.4,
  "SC 0.5 mg/kg",   0.5931,
  "SC 1.7 mg/kg",   4.005,
  "SC 3.3 mg/kg",  18.52
) |>
  dplyr::mutate(auclast = auclast_ugmL / nM_to_ugmL) |>
  dplyr::select(arm, auclast)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by        = "arm",
  params    = "auclast",
  units     = c(auclast = "day*nM"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste("Simulated vs Kwak 2025 Supplementary Table S3 AUC(0-30 days).",
                  "* marks a >20% difference."),
  align = c("l", "l", "r", "r", "r")
)
```

```{r nca-gate}
# Independent numeric gate on the same six values (the table above is for the
# reader; this is what fails the render). Deterministic model, exactly specified
# window -- so the only slack needed is trapezoidal-integration error on the
# steep distribution phase plus the paper's 4-significant-figure rounding.
gate <- human_single |>
  dplyr::mutate(arm = paste(route, dose)) |>
  dplyr::group_by(arm) |>
  dplyr::group_modify(~ tibble::tibble(
    sim_ugmL = auc_of(.x, "Cc", 0, 30) * nM_to_ugmL)) |>
  dplyr::ungroup() |>
  dplyr::left_join(
    published |> dplyr::mutate(paper_ugmL = auclast * nM_to_ugmL) |>
      dplyr::select(arm, paper_ugmL),
    by = "arm") |>
  dplyr::mutate(`% diff` = 100 * (sim_ugmL - paper_ugmL) / paper_ugmL)

gate |>
  dplyr::mutate(dplyr::across(c(sim_ugmL, paper_ugmL), ~ signif(.x, 4)),
                `% diff` = round(`% diff`, 2)) |>
  dplyr::rename("Arm" = arm,
                "Simulated AUC0-30d (day*ug/mL)" = sim_ugmL,
                "Table S3 (day*ug/mL)" = paper_ugmL) |>
  knitr::kable(caption = "AUC(0-30 days): simulation vs the paper's own predicted values.")

# Realised max |% diff| ~1.2% on the grid above. 5% still goes red on a
# mis-transcribed rate constant, volume or dose conversion, all of which move
# these AUCs by tens of percent.
stopifnot(max(abs(gate$`% diff`)) < 5)
```

The paper also reports that dose-normalised AUC rises **more than
proportionally** with dose, more steeply after s.c. than i.v. dosing -- the
signature of saturable target binding (section 3.3, Supplementary Table S3).

```{r dose-proportionality}
dn <- gate |>
  tidyr::separate(arm, into = c("route", "dose_mgkg"), sep = " ", extra = "merge") |>
  dplyr::mutate(dose_mgkg = as.numeric(sub(" mg/kg", "", dose_mgkg)),
                dn_auc = paper_ugmL / dose_mgkg) |>
  dplyr::arrange(route, dose_mgkg)

dn |>
  dplyr::mutate(dn_auc = signif(dn_auc, 4)) |>
  dplyr::select(Route = route, `Dose (mg/kg)` = dose_mgkg,
                `Dose-normalised AUC0-30d` = dn_auc) |>
  knitr::kable(caption = "Dose-normalised AUC increases with dose (more-than-proportional exposure).")

fold <- dn |>
  dplyr::group_by(route) |>
  dplyr::summarise(fold = dn_auc[which.max(dose_mgkg)] / dn_auc[which.min(dose_mgkg)],
                   .groups = "drop")
print(fold)

stopifnot(
  # More than dose proportional in BOTH routes ...
  all(fold$fold > 1),
  # ... and more markedly so after s.c. dosing (section 3.3).
  fold$fold[fold$route == "SC"] > fold$fold[fold$route == "IV"]
)
```

# The paper's dosing conclusions, checked

Section 3.3 makes four specific, deterministic claims about human sTFPI-alpha
suppression. Each is evaluated against the simulation.

```{r claims}
thresh <- 0.25 * stfpi_base_human

# Rows of one arm, optionally restricted to a time window, sorted by time.
arm_rows <- function(df, dose_label, route_label, from = 0, to = Inf) {
  df |>
    dplyr::filter(dose == dose_label, route == route_label,
                  time >= from, time <= to) |>
    dplyr::arrange(time)
}
# Does the arm ever get below 25% of baseline?
ever_below <- function(...) any(arm_rows(...)$sTFPI < thresh)
# Time (days) spent below 25% of baseline within the window.
dur_below <- function(...) {
  d <- arm_rows(...)
  # Guard against a lookup that silently matches no rows (pattern 10).
  stopifnot(nrow(d) > 1)
  sum(diff(d$time) * utils::head(d$sTFPI < thresh, -1))
}

iv_all   <- vapply(names(human_doses_nmol),
                   function(d) ever_below(human_single, d, "IV"), logical(1))
sc_any   <- vapply(names(human_doses_nmol),
                   function(d) ever_below(human_single, d, "SC"), logical(1))
sc33_dur <- dur_below(human_single, "3.3 mg/kg", "SC")
# Final full dosing interval of the Q7d run (doses at 0..49; interval 49-56).
ss_frac  <- dur_below(human_q7d, "3.3 mg/kg", "SC", from = 49, to = 56) / 7

claims <- tibble::tribble(
  ~Claim, ~Achieved, ~Pass,
  "Single i.v. drives sTFPI-a below 25% of baseline at ALL doses",
  paste(sprintf("%s: %s", names(iv_all), iv_all), collapse = "; "),
  all(iv_all),

  "Single s.c. reaches it ONLY at 3.3 mg/kg",
  paste(sprintf("%s: %s", names(sc_any), sc_any), collapse = "; "),
  identical(unname(sc_any), c(FALSE, FALSE, TRUE)),

  "Single s.c. 3.3 mg/kg holds suppression for ~3 days",
  sprintf("%.2f days", sc33_dur),
  abs(sc33_dur - 3) < 1.5,

  "Q7d s.c. 3.3 mg/kg holds it for the MAJORITY of the steady-state interval",
  sprintf("%.0f%% of the final 7-day interval", 100 * ss_frac),
  ss_frac > 0.5
)

knitr::kable(claims, caption = "Kwak 2025 section 3.3 dosing conclusions, evaluated against the packaged human model.")

stopifnot(all(claims$Pass))
```

# Assumptions and deviations

* **Publication year.** The model files, the vignette basename and the citation
  use **2025**, matching the DOI (`10.3389/fphar.2025.1745702`), the EuropePMC
  `pubYear`, and the PMCID record. The journal's own suggested-citation block
  reads "(2026)" because Frontiers first published the article online on
  2026-01-07 against a 2025 DOI. No content differs between the two.
* **ODEs rewritten from concentration to amount for two states.** The paper
  writes `dC2/dt` and `dC3/dt` on concentration; the model files carry `central`
  and `peripheral1` as amounts (each equation multiplied through by its volume)
  so that i.v. and s.c. dosing both work through a standard rxode2 event table.
  This is an algebraically identical restatement, not a modelling change. The
  four TFPI-species states remain concentrations in nM as published.
* **No IIV and no residual error are encoded, because none are reported.** CGNM
  was fitted to the *mean* observed profiles by least squares; the paper reports
  no random-effects model of any kind. These files are typical-value
  deterministic simulators and cannot be used for individual-level VPCs without
  variability supplied from elsewhere.
* **Five parameters are non-identifiable.** Approximate profile likelihood
  (Supplementary Table S1) declares `k02`, `kel,CM`, `kel,CS`, `kloss` and `kon`
  non-identifiable. `kon` had two local minima; the authors used the first, which
  is what is encoded. `k02` and `kloss` are retained at their minute rank 1
  values rather than set to zero, following the authors' explicit choice to
  "maintain a general model applicable to other antibody therapeutics"
  (section 3.1); Supplementary Table S2 shows fixing them to zero changes the
  remaining estimates and the minimum SSR negligibly. Do not read these five
  values as well-determined.
* **The rabbit and human files are predictions, not fits.** Every parameter in
  them is wrapped in `fixed()` for that reason. The rabbit projection was
  evaluated against observed data and over-predicts the low-dose arms (AAFE
  2.8-4.2); the human projection was **never validated against any human data**,
  which the authors state explicitly. Treat the human file as a forward
  prediction under a stated allometric assumption.
* **Binding parameters assumed conserved across species.** `KD` and `kon` are not
  scaled, on the strength of K2-domain sequence similarity (human vs monkey 96%
  identity; human vs rabbit 92%). The fitted in-vivo `kon` (28.5 1/(nM*day)) is
  roughly 10-fold below the in-vitro value of Kwak 2020; the authors keep the
  in-vitro `KD` and let `koff` follow (section 4).
* **Monkey Table 2 AUC(last) is not reproduced numerically.** Table 2 compares
  observed and predicted AUC to the *last measurable concentration*, but neither
  the sampling schedule nor the per-arm last measurable time is tabulated in this
  paper, so the integration window cannot be reconstructed. The exactly specified
  Supplementary Table S3 AUC(0-30 days) is used as the numeric gate instead.
* **Rabbit observed data are not tabulated** in the paper (Figure 3 only), so the
  published AAFE values are recorded above rather than recomputed.
* **`sTFPI-alpha` baseline as a source of variability.** The paper explores a
  1.3-2.9 nM human baseline by local sensitivity analysis rather than as a fitted
  covariate; it is recorded in the human file's `covariatesDataExcluded` metadata.
  To reproduce Figure 5, set `stfpi_b` and rescale `mtfpi_b = 12.7 * stfpi_b`.
* **Supplement provenance.** The model equations are in the Supplementary
  Material (`DataSheet1.pdf`), retrieved from the EuropePMC supplementary-files
  endpoint for PMC12819659 and stored alongside the lead PDF. They are not in the
  main article text.
