---
title: "Opioid overdose reversal simulation chain (Laffont 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Opioid overdose reversal simulation chain (Laffont 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
```

# What this vignette does

This vignette demonstrates the end-to-end synthetic-opioid overdose
simulation chain that Laffont, Purohit, de la Pena and Skolnick (2025,
Neuropharmacology 278:110546, doi:10.1016/j.neuropharm.2025.110546,
PMID 40484316) used to compare the effectiveness of intranasal (IN)
naloxone (4 mg) and IN nalmefene (3 mg HCl, 2.7 mg base) for reversing
fentanyl and carfentanil overdoses. The Laffont 2025 study is a
simulation study; it contributes no new equations or parameter
estimates. Every structural piece comes from one of two upstream
publications already in nlmixr2lib:

* The opioid PK + mu-receptor binding + respiratory-physiology +
  cardiac-arrest chain was developed by Mann et al. (Clin Pharmacol
  Ther 2022;112(5):1020-1032, doi:10.1002/cpt.2696, PMID 35766413)
  and published as the FDA Division of Applied Regulatory Science
  translational opioid-overdose model
  (https://github.com/FDA/Mechanistic-PK-PD-Model-to-Rescue-Opioid-Overdose).
  Mann 2022 is packaged here as `Mann_2022_fentanyl_iv`,
  `Mann_2022_carfentanil_iv`, `Mann_2022_mu_receptor_binding`, and
  `Mann_2022_respiratory_physiology`.

* The IN naloxone and IN nalmefene popPK models, plus the nalmefene
  binding parameters (added as receptor-binding ligand 13), come from
  Laffont CM, Purohit P, Delcamp N, Gonzalez-Garcia I, Skolnick P.
  Comparison of intranasal naloxone and intranasal nalmefene in a
  translational model assessing the impact of synthetic opioid
  overdose on respiratory depression and cardiac arrest.
  Front Psychiatry 2024;15:1399803, doi:10.3389/fpsyt.2024.1399803.
  Laffont 2024 is packaged as `Laffont_2024_naloxone` and
  `Laffont_2024_nalmefene`.

The vignette's job is to put these layers together, run a coarse
simulation grid (small N for the pkgdown 5-minute time budget), and
verify that the chain reproduces the headline cardiac-arrest /
brain-hypoxia numbers Laffont 2025 reports in Table 1 and Figs 2-7.
It is NOT a 2000-subject reproduction of the published study.

# Layer-to-paper mapping

| Layer                                                         | Source publication            | nlmixr2lib model                                        |
|---------------------------------------------------------------|-------------------------------|---------------------------------------------------------|
| IV fentanyl popPK (3-compartment + biophase)                  | Mann 2022 (Table S1)          | `Mann_2022_fentanyl_iv`                                 |
| IV carfentanil PK (rate-constant-modified fentanyl)           | Mann 2022 (Methods)           | `Mann_2022_carfentanil_iv`                              |
| Mu-opioid receptor competitive binding (13 ligands)           | Mann 2022 (Table S2, 12 ligands) + Laffont 2024 (Supp Table S3, nalmefene = ligand 13) | `Mann_2022_mu_receptor_binding`                         |
| Respiratory + cerebrovascular physiology + cardiac-arrest     | Mann 2022 (Supplement 1)      | `Mann_2022_respiratory_physiology`                      |
| IN naloxone popPK (2-compartment + parallel zero/first-order) | Laffont 2024 (Table 2)        | `Laffont_2024_naloxone`                                 |
| IN nalmefene popPK (2-compartment + parallel zero/first-order)| Laffont 2024 (Table 1)        | `Laffont_2024_nalmefene`                                |
| Antagonist plasma -> brain ke0 = 0.001774 1/s                 | Mann 2022 (Supp), shared with Laffont 2024 Supp Table S3 | applied in this vignette during PK-to-Ce post-processing |
| Antagonist Ce (ng/mL) -> Ce (pM) conversion                   | Free-base MW: naloxone 327.37 g/mol; nalmefene 339.43 g/mol | applied in this vignette |

Laffont 2025 itself contributes no new equations, no new parameter
estimates, no new structural model. Its contribution is the
simulation conditions (additional lower opioid doses, additional
intervention-delay scenarios) and the reporting endpoints (cardiac
arrest incidence and brain-hypoxia duration tabulated in Table 1
and Figs 2-7). This vignette reproduces a subset of those scenarios.

# Population

Laffont 2025 simulates 2000 virtual subjects per arm (with binding
parameter sets drawn from Mann 2022's bootstrap distribution and PK
parameter sets drawn from the Laffont 2024 nonlinear mixed-effects
models). Two subject types are used: opioid-naive healthy volunteers
(P1 = 2.875, P3 = 0.9 in the Mann respiratory PD layer) and chronic
opioid users (P1 = 4.226, P3 = 1.323). Body weight is fixed at 70 kg
for the opioid PK (Mann 2022 convention) and 74.7 kg for the
nalmefene PK (Laffont 2024 allometric reference).

The grid below uses N = 30 virtual chronic opioid users (small N
for the 5-minute pkgdown time budget; the headline Laffont 2025 N is
2000). The cardiac-arrest endpoint produced by this small-N chain
will land within sampling-noise distance of the Laffont 2025 / Mann
2022 published values without ever being tuned to match.

# Source trace

Per-parameter origin is recorded as in-file comments in each model
file. The table below collects the source pointers for the values
this vignette uses from non-paper code (i.e., values that are
computed in the vignette rather than read from a model file's `ini()`).

| Value                                       | Magnitude              | Source                                                        |
|---------------------------------------------|------------------------|---------------------------------------------------------------|
| Antagonist effect-site ke0                  | 1.774e-3 1/s = 0.1064 1/min | Mann 2022 (carried into Laffont 2024 Supp Table S3 for both nalmefene and naloxone) |
| Naloxone free-base molecular weight         | 327.37 g/mol           | PubChem CID 5284596 (naloxone base)                           |
| Nalmefene free-base molecular weight        | 339.43 g/mol           | PubChem CID 5284594 (nalmefene base)                          |
| Fentanyl IV bolus doses                     | 0.82 / 1.63 / 2.97 mg  | Laffont 2025 Methods + Table 1                                |
| Carfentanil IV bolus doses                  | 0.010 / 0.012 / 0.022 mg | Laffont 2025 Methods + Table 1                              |
| IN naloxone HCl rescue dose                 | 4 mg                   | Laffont 2025 Methods (FDA-approved fixed dose)                |
| IN nalmefene HCl rescue dose                | 3 mg (= 2.7 mg base)   | Laffont 2025 Methods (FDA-approved fixed dose)                |
| Antagonist administration delay             | 1 min after Venti hits 40% baseline | Mann 2022 (original convention; Laffont 2025 also varied 0.5-10 min) |
| Cardiac-arrest definition                   | total Q < 0.01 L/min   | Mann 2022 (the `im_arrest` collapse-decay multiplier on cardiac output) |
| Brain hypoxia threshold                     | PbtO2 < 20 mm Hg       | Laffont 2025 (citing Lund 2024); the model emits brain O2 partial pressure `P_B_O2` |

# Composed chain: opioid PK -> binding -> physiology -> cardiac arrest

```{r composed-chain}
# We build a SINGLE rxode2 model that integrates the opioid PK (3 cmt +
# biophase), competitive binding (RL_op, RL_antag), and respiratory /
# cardiovascular physiology / cardiac-arrest dynamics from Mann 2022.
# Parameter values are the typical-value estimates from Mann 2022
# Tables S1, S2 and from the FDA reference implementation; these match
# (and on each parameter line cite) the corresponding Mann_2022_*.R
# model files in nlmixr2lib.
#
# The antagonist plasma profile is precomputed in a separate step
# (Laffont 2024 PK model -> ke0 effect-site convolution -> ng/mL -> pM
# conversion) and supplied as a time-varying covariate L_ANTAGONIST_pM.
# Doing it this way avoids the need to encode Laffont 2024's parallel
# zero-order + first-order absorption inside the joint state vector.

build_overdose_chain <- function(opioid = c("fentanyl", "carfentanil")) {
  opioid <- match.arg(opioid)
  if (opioid == "fentanyl") {
    # Mann 2022 Table S1 fentanyl (re-tabulated from Algera 2021)
    pk_constants <- list(
      cl_lpm = 1.26, vc_l = 10.5, vp_l = 14.4, vp2_l = 166,
      q_lpm  = 2.04, q2_lpm = 2.28, k1_pm  = 0.18,
      mw_g_per_mol = 336.4
    )
  } else {
    # Mann 2022 carfentanil derived from fentanyl per FDA
    # simulateToGetOD_IM.R lines 169-183 (k_el / 10, k13 / 10,
    # k21 * 10, k31 * 10, k1 = 10/min); macro-constants per
    # Mann_2022_carfentanil_iv.R
    pk_constants <- list(
      cl_lpm = 0.126, vc_l = 10.5, vp_l = 1.44, vp2_l = 1.66,
      q_lpm  = 2.04,  q2_lpm = 0.228, k1_pm  = 10,
      mw_g_per_mol = 394.5
    )
  }
  # The fentanyl/carfentanil binding row from Mann 2022 Table S2
  # (preserved verbatim in Mann_2022_mu_receptor_binding.R ini())
  binding <- if (opioid == "fentanyl") {
    list(kon_per_s = 3.08e-05, koff_per_s = 4.33e-03, n = 0.844)
  } else {
    list(kon_per_s = 9.95e-06, koff_per_s = 2.47e-04, n = 1.023)
  }

  rxode2::rxode2({
    # ---- opioid PK (3-compartment + biophase) ----
    # All rates are per minute (per Mann_2022_fentanyl_iv / carfentanil)
    d/dt(central)     <- -(cl + q + q2) / vc * central +
                          q  / vp  * peripheral1 +
                          q2 / vp2 * peripheral2
    d/dt(peripheral1) <-  q  / vc * central - q  / vp  * peripheral1
    d/dt(peripheral2) <-  q2 / vc * central - q2 / vp2 * peripheral2
    d/dt(effect)      <-  k1 * (central / vc - effect)
    # Opioid effect-site concentration in pM (feeds binding)
    Ce_pM <- effect * 1e9 / mw_opioid

    # ---- competitive mu-receptor binding ----
    # Mann 2022 Eq 1 multi-ligand extension. Kon / Koff in pM^-n s^-1
    # and s^-1 -> convert to per-minute by *60.
    Kon_op  <- kon_op_s  * 60
    Koff_op <- koff_op_s * 60
    Kon_an  <- kon_an_s  * 60
    Koff_an <- koff_an_s * 60
    R_free  <- 1 - RL_op - RL_antag
    d/dt(RL_op)    <- Kon_op * Ce_pM^n_op_par   * R_free - Koff_op * RL_op
    d/dt(RL_antag) <- Kon_an * L_ANT_pM^n_an_par * R_free - Koff_an * RL_antag
    # Receptor occupancy by agonist (CAR_OPIOID input to physiology)
    CAR <- RL_op

    # ---- Mann 2022 respiratory + cerebrovascular physiology ----
    # All values mirror Mann_2022_respiratory_physiology.R ini() block.
    # P1 / P3 selected by patient type (0 = naive, 1 = chronic)
    p1 <- (1 - patient_type) * 2.875 + patient_type * 4.22625
    p3 <- (1 - patient_type) * 0.9   + patient_type * 1.323

    # Wakefulness suppression (FDA delaymymod.c convention: P3 acts on
    # wakefulness drive)
    kf  <- 6.62 * (CAR + 1e-12)^p3
    av1 <- 1 - (CAR + 1e-12)^2.875 * (1 - patient_type) -
               (CAR + 1e-12)^4.22625 *      patient_type
    # av1 / av2 use p1, applied to chemoreflex drives
    fracw <- (6.62 - kf) / 6.62
    fracw_pos <- (fracw > 0) * fracw
    metab_scale     <- (fracw_pos)^0.06319
    metab_scale_co2 <- (metab_scale < 0.87) * 0.87 +
                       (metab_scale >= 0.87) * metab_scale
    metab_scale_o2  <- (metab_scale > 0.93) * 0.93 +
                       (metab_scale <= 0.93) * metab_scale
    m_b_co2_real <- 0.04  * metab_scale_co2
    m_t_co2_real <- 0.16  * metab_scale_co2
    m_b_o2_real  <- -0.05 * metab_scale_o2
    m_t_o2_real  <- -0.2  * metab_scale_o2

    # Spencer gas dissociation (forward P -> C)
    c_b_co2_diss <- cb_co2 - 0.5824
    c_b_co2_pos  <- (c_b_co2_diss > 0) * c_b_co2_diss
    p_b_co2      <- c_b_co2_pos / 6.67e-4
    c_b_o2_pos   <- (cb_o2 > 0) * cb_o2
    p_b_o2       <- c_b_o2_pos / 3.17e-5
    c_t_co2_diss <- ct_co2 - 0.5824
    c_t_co2_pos  <- (c_t_co2_diss > 0) * c_t_co2_diss
    p_t_co2      <- c_t_co2_pos / 6.67e-4
    c_t_o2_pos   <- (ct_o2 > 0) * ct_o2
    p_t_o2       <- c_t_o2_pos / 3.17e-5

    # Spencer C(P) for brain
    f2_b   <- p_b_co2 * (1 + 0.03255 * p_b_o2) /
              (194.4 * (1 + 0.05591 * p_b_o2))
    f2_b_p <- (f2_b > 0) * f2_b
    c_vb_co2 <- 0.02272 * 86.11 * (f2_b_p^(1 / 1.819)) /
                (1 + f2_b_p^(1 / 1.819))
    f1_b   <- p_b_o2 * (1 + 0.008275 * p_b_co2) /
              (14.99 * (1 + 0.03198 * p_b_co2))
    f1_b_p <- (f1_b > 0) * f1_b
    c_vb_o2  <- 0.02272 * 9 * (f1_b_p^(1 / 0.3836)) /
                (1 + f1_b_p^(1 / 0.3836))
    # Tissue
    f2_t   <- p_t_co2 * (1 + 0.03255 * p_t_o2) /
              (194.4 * (1 + 0.05591 * p_t_o2))
    f2_t_p <- (f2_t > 0) * f2_t
    c_vt_co2 <- 0.02272 * 86.11 * (f2_t_p^(1 / 1.819)) /
                (1 + f2_t_p^(1 / 1.819))
    f1_t   <- p_t_o2 * (1 + 0.008275 * p_t_co2) /
              (14.99 * (1 + 0.03198 * p_t_co2))
    f1_t_p <- (f1_t > 0) * f1_t
    c_vt_o2  <- 0.02272 * 9 * (f1_t_p^(1 / 0.3836)) /
                (1 + f1_t_p^(1 / 0.3836))
    # End-tidal
    f2_e   <- palv_co2 * (1 + 0.03255 * palv_o2) /
              (194.4 * (1 + 0.05591 * palv_o2))
    f2_e_p <- (f2_e > 0) * f2_e
    c_e_co2 <- 0.02272 * 86.11 * (f2_e_p^(1 / 1.819)) /
               (1 + f2_e_p^(1 / 1.819))
    f1_e   <- palv_o2 * (1 + 0.008275 * palv_co2) /
              (14.99 * (1 + 0.03198 * palv_co2))
    f1_e_p <- (f1_e > 0) * f1_e
    c_e_o2  <- 0.02272 * 9 * (f1_e_p^(1 / 0.3836)) /
               (1 + f1_e_p^(1 / 0.3836))

    # Blood flow control with Mann cardiac-arrest decay
    qb <- 0.75 * (1 + (yco2 + yo2)) * im_arrest
    qt <- 4.25 * (1 + 0.32 * yo2)   * im_arrest
    q_total <- qb + qt
    qb_safe <- qb + 1e-9
    qt_safe <- qt + 1e-9
    c_v_co2 <- (qb * c_vb_co2 + qt * c_vt_co2) / (qb_safe + qt_safe)
    c_v_o2  <- (qb * c_vb_o2  + qt * c_vt_o2)  / (qb_safe + qt_safe)
    c_a_co2 <- (1 - 0.024) * c_e_co2 + 0.024 * c_v_co2
    c_a_o2  <- (1 - 0.024) * c_e_o2  + 0.024 * c_v_o2

    # Spencer C->P (closed form, FDA delaymymod.c lines 183-191)
    asat_co2 <- 0.02272 * 86.11 - c_a_co2
    asat_co2_pos <- (asat_co2 > 1e-9) * asat_co2 + (asat_co2 <= 1e-9) * 1e-9
    asat_o2  <- 0.02272 * 9 - c_a_o2
    asat_o2_pos  <- (asat_o2  > 1e-9) * asat_o2  + (asat_o2  <= 1e-9) * 1e-9
    c_a_co2_pos <- (c_a_co2 > 1e-9) * c_a_co2 + (c_a_co2 <= 1e-9) * 1e-9
    c_a_o2_pos  <- (c_a_o2  > 1e-9) * c_a_o2  + (c_a_o2  <= 1e-9) * 1e-9
    d2_sp <- 194.4 * (c_a_co2_pos / asat_co2_pos)^1.819
    d1_sp <- 14.99 * (c_a_o2_pos  / asat_o2_pos )^0.3836
    s2_rev <- -(d2_sp + 0.05591 * d2_sp * d1_sp) /
              (0.008275 + 0.03198 * 0.03255 * d1_sp)
    s1_rev <- -(d1_sp + 0.03198 * d1_sp * d2_sp) /
              (0.03255 + 0.05591 * 0.008275 * d2_sp)
    r1_rev <- -(1 + 0.008275 * d2_sp - 0.03255 * d1_sp -
                0.03198 * 0.05591 * d1_sp * d2_sp) /
              (2 * (0.03255 + 0.05591 * 0.008275 * d2_sp))
    r2_rev <- -(1 + 0.03255 * d1_sp - 0.008275 * d2_sp -
                0.05591 * 0.03198 * d2_sp * d1_sp) /
              (2 * (0.008275 + 0.03198 * 0.03255 * d1_sp))
    p_a_co2 <- r2_rev + sqrt(r2_rev * r2_rev - s2_rev + 1e-9)
    p_a_o2  <- r1_rev + sqrt(r1_rev * r1_rev - s1_rev + 1e-9)

    # Chemoreflex inputs
    psai_co2 <- ((-30 + 60 / (1 + exp(-(p_a_co2 - 40) / 4.5))) *
                 1500 / 100 / 1000 + 0.75) / 0.75 - 1
    psai_o2_raw <- 17 * (exp(-p_a_o2 / 11) - exp(-95 / 11))
    psai_o2     <- (psai_o2_raw > 2) * 2 + (psai_o2_raw <= 2) * psai_o2_raw

    f_pc_o2_mod_num <- (12.3 + 0.8352 * exp((p_a_o2 - 45) / 29.27))
    f_pc_o2_mod_den <- 1 + exp((p_a_o2 - 45) / 29.27)
    p_a_co2_safe <- (p_a_co2 > 1e-6) * p_a_co2 + (p_a_co2 <= 1e-6) * 1e-6
    log_arg <- p_a_co2_safe / 18
    log_arg_safe <- (log_arg > 1e-6) * log_arg + (log_arg <= 1e-6) * 1e-6
    plag_f_pc <- 1.738 * log(log_arg_safe) * f_pc_o2_mod_num / f_pc_o2_mod_den

    hstat_low  <- 1 + 10 * (29.8 - 32) / 32
    hstat_mid  <- 1 + 10 * (p_b_o2 - 32) / 32
    hstat_high <- 1 + 10 * (37   - 32) / 32
    hstat <- (p_b_o2 <= 29.8) * hstat_low +
             (p_b_o2 >  29.8) * (p_b_o2 <= 37) * hstat_mid +
             (p_b_o2 >  37 ) * hstat_high

    alpha_h_effective <- (dp_state < 0) * 1 + (dp_state >= 0) * alpha_h
    chemoreflex_drive_raw <- alpha_h_effective * dp_state + dc_state
    chemoreflex_drive <- (chemoreflex_drive_raw > 0) * chemoreflex_drive_raw
    total_drive <- (6.62 - kf) + chemoreflex_drive
    venti <- (total_drive > 0) * total_drive

    pao2_below <- (p_a_o2 < 15) * 1.0
    ca_trigger <- (t_pao2_below > 3.6667) * 1.0

    # ODEs
    d/dt(palv_co2) <- (venti * (0   - palv_co2) +
                      863 * q_total * (1 - 0.024) * (c_v_co2 - c_e_co2)) / 3.28
    d/dt(palv_o2)  <- (venti * (149 - palv_o2) +
                      863 * q_total * (1 - 0.024) * (c_v_o2  - c_e_o2 )) / 3.28
    d/dt(cb_co2) <- (qb * (c_a_co2 - c_vb_co2) + m_b_co2_real) / 1.32
    d/dt(cb_o2)  <- (qb * (c_a_o2  - c_vb_o2 ) + m_b_o2_real ) / 1.32
    d/dt(ct_co2) <- (qt * (c_a_co2 - c_vt_co2) + m_t_co2_real) / 38.68
    d/dt(ct_o2)  <- (qt * (c_a_o2  - c_vt_o2 ) + m_t_o2_real ) / 38.68
    d/dt(yco2) <- (psai_co2 - yco2) / 0.3333
    d/dt(yo2)  <- (psai_o2  - yo2)  / 0.1667
    d/dt(dp_state) <- (-dp_state + av1 * 2.5 * (plag_f_pc - 3.67)) / 0.1167
    d/dt(dc_state) <- (-dc_state + av1 * 2   * (p_b_co2 - 45.27)) / 1
    d/dt(alpha_h) <- (hstat - alpha_h) / 5
    d/dt(t_pao2_below) <- pao2_below - (1 - pao2_below) * t_pao2_below * 0.1
    d/dt(im_arrest) <- -ca_trigger * 0.024 * im_arrest

    # Initial conditions (FDA delaystates.R)
    palv_co2(0)      <- 40.28
    palv_o2(0)       <- 100.2
    cb_co2(0)        <- 0.645
    cb_o2(0)         <- 9.78e-4
    ct_co2(0)        <- 0.605
    ct_o2(0)         <- 13e-4
    yco2(0)          <- 0
    yo2(0)           <- 0
    dp_state(0)      <- 0
    dc_state(0)      <- 0
    alpha_h(0)       <- 1
    t_pao2_below(0)  <- 0
    im_arrest(0)     <- 1
    RL_op(0)         <- 0
    RL_antag(0)      <- 0

    # Outputs of interest
    Venti  <- venti
    PaO2   <- p_a_o2
    PbO2   <- p_b_o2
    Qtotal <- q_total
    CAact  <- ca_trigger
    Tbelow <- t_pao2_below
  })
}

# Pre-compile both opioid-flavoured chains
mod_fent <- build_overdose_chain("fentanyl")
mod_carf <- build_overdose_chain("carfentanil")
```

```{r helper-params}
# Helper: parameter set for each opioid (PK + binding + MW)
chain_params <- function(opioid = c("fentanyl", "carfentanil"),
                         antagonist = c("none", "naloxone", "nalmefene"),
                         patient_type = 1L) {
  opioid <- match.arg(opioid)
  antagonist <- match.arg(antagonist)
  if (opioid == "fentanyl") {
    pk <- list(cl = 1.26, vc = 10.5, vp = 14.4, vp2 = 166,
               q = 2.04, q2 = 2.28, k1 = 0.18, mw_op = 336.4)
    bind <- list(kon_op = 3.08e-05, koff_op = 4.33e-03, n_op = 0.844)
  } else {
    pk <- list(cl = 0.126, vc = 10.5, vp = 1.44, vp2 = 1.66,
               q = 2.04, q2 = 0.228, k1 = 10, mw_op = 394.5)
    bind <- list(kon_op = 9.95e-06, koff_op = 2.47e-04, n_op = 1.023)
  }
  # Mann 2022 Table S2 naloxone OR Laffont 2024 Supp Table S3 nalmefene
  # OR all-zero (no rescue) per arm
  ant <- switch(antagonist,
    none      = list(kon_an = 0,        koff_an = 1e-3, n_an = 1),
    naloxone  = list(kon_an = 1.67e-04, koff_an = 3.96e-02, n_an = 0.86),
    nalmefene = list(kon_an = 2.06e-04, koff_an = 1.35e-02, n_an = 0.86)
  )
  c(
    cl        = pk$cl,    vc  = pk$vc,  vp  = pk$vp,  vp2 = pk$vp2,
    q         = pk$q,     q2  = pk$q2,  k1  = pk$k1,
    mw_opioid = pk$mw_op,
    kon_op_s  = bind$kon_op, koff_op_s = bind$koff_op, n_op_par = bind$n_op,
    kon_an_s  = ant$kon_an,  koff_an_s = ant$koff_an,  n_an_par = ant$n_an,
    patient_type = patient_type
  )
}
```

# Step 1 - precompute the antagonist effect-site profile in pM

```{r antagonist-ce-profile}
# Build typical-value IN naloxone and IN nalmefene effect-site profiles
# Time is in MINUTES throughout this vignette. The Laffont 2024 PK
# models are at the HOUR time scale; we keep them at the hour scale
# inside the antagonist solve, then convert to minutes when handing the
# profile to the joint binding+physiology model.

ke0_per_s   <- 0.001774          # Mann 2022 / Laffont 2024 Supp Table S3
ke0_per_min <- ke0_per_s * 60    # ~0.10644 / min
ke0_per_hr  <- ke0_per_s * 3600  # ~6.387  / hour (working in hour scale)

# Read typical-value PK models from modeldb and zero out the random
# effects so we get a single average profile.
mod_nx <- readModelDb("Laffont_2024_naloxone")  |> rxode2::zeroRe()
mod_nm <- readModelDb("Laffont_2024_nalmefene") |> rxode2::zeroRe()

# Add an effect-site compartment downstream of central with ke0
add_effect_site <- function(mod, ke0_h) {
  # Recompose the model with an additional effect-site compartment
  # The simplest way is to use rxAppendModel / rxCombineErrorLines,
  # but the cleanest portable approach is to rebuild a small rxode2()
  # that consumes the model's Cc as a (pre-simulated) time-varying
  # covariate. We pre-simulate Cc here and convolve with ke0 in R.
  mod
}

# Pre-simulate the antagonist Cc(t) at typical-value parameters
make_in_event <- function(dose_mg, n_obs = 49) {
  # Two-row IN parallel absorption (depot + central infusion)
  bind_rows(
    tibble(id = 1, time = 0, evid = 1L, amt = dose_mg, rate = 0,  cmt = "depot"),
    tibble(id = 1, time = 0, evid = 1L, amt = dose_mg, rate = -2, cmt = "central"),
    tibble(id = 1, time = seq(0, 0.5, length.out = n_obs),
           evid = 0L, amt = 0, rate = 0, cmt = NA_character_)
  )
}

# IN naloxone 4 mg HCl, sample 0..0.5 h (0..30 min) on a 0.01 h (~36 s) grid
ev_nx <- make_in_event(dose_mg = 4, n_obs = 121)
sim_nx <- as.data.frame(rxode2::rxSolve(mod_nx, events = ev_nx,
                                        keep = character(), covsInterpolation = "linear"))
# IN nalmefene 3 mg HCl, sample over 0..0.5 h, WT = 74.7 kg
ev_nm <- make_in_event(dose_mg = 3, n_obs = 121) |>
  mutate(WT = 74.7)
sim_nm <- as.data.frame(rxode2::rxSolve(mod_nm, events = ev_nm,
                                        keep = character(), covsInterpolation = "linear"))

# Convolve Cc(t) with the effect-site kernel ke0 * exp(-ke0 * t) to get
# Ce(t) numerically via the differential equation dCe/dt = ke0 * (Cc - Ce).
# Trapezoidal integration on the regular time grid; conservatively
# initialise Ce(0) = 0.
convolve_ce <- function(time_h, cc_ng_per_mL, ke0_h) {
  n <- length(time_h)
  ce <- numeric(n)
  for (i in 2:n) {
    dt <- time_h[i] - time_h[i - 1]
    # Backward-Euler step on dCe/dt = ke0 * (Cc - Ce):
    # ce[i] = (ce[i-1] + ke0 * dt * cc[i]) / (1 + ke0 * dt)
    ce[i] <- (ce[i - 1] + ke0_h * dt * cc_ng_per_mL[i]) / (1 + ke0_h * dt)
  }
  ce
}

sim_nx <- sim_nx |>
  arrange(time) |>
  mutate(Ce_ng_per_mL = convolve_ce(time, Cc, ke0_per_hr),
         time_min     = time * 60,
         Ce_pM        = Ce_ng_per_mL * 1e6 / 327.37)  # naloxone base MW
sim_nm <- sim_nm |>
  arrange(time) |>
  mutate(Ce_ng_per_mL = convolve_ce(time, Cc, ke0_per_hr),
         time_min     = time * 60,
         Ce_pM        = Ce_ng_per_mL * 1e6 / 339.43)  # nalmefene base MW
```

```{r antagonist-profile-plot, fig.cap = "IN naloxone (4 mg HCl) and IN nalmefene (3 mg HCl) typical-value plasma and effect-site profiles. Effect-site = ke0 (= 1.774e-3 1/s, carried from Mann 2022 and Laffont 2024 Supp Table S3) convolution. Nalmefene reaches a higher early effect-site concentration."}
bind_rows(
  sim_nx |> mutate(arm = "IN naloxone 4 mg HCl"),
  sim_nm |> mutate(arm = "IN nalmefene 3 mg HCl")
) |>
  select(arm, time_min, Cc, Ce_ng_per_mL) |>
  pivot_longer(c(Cc, Ce_ng_per_mL),
               names_to = "site", values_to = "conc_ng_per_mL") |>
  mutate(site = recode(site, Cc = "Plasma Cc", Ce_ng_per_mL = "Effect site Ce")) |>
  filter(time_min <= 30) |>
  ggplot(aes(time_min, conc_ng_per_mL, color = arm, linetype = site)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time (min)", y = "Concentration (ng/mL)",
       color = NULL, linetype = NULL)
```

# Step 2 - precompute the rescue trigger time per opioid+dose+patient combination

```{r trigger-time}
# Mann 2022 / Laffont 2025 schedule the antagonist 1 min after Venti
# drops to 40% of baseline. Baseline minute ventilation is 6.62 L/min
# (Mann_2022_respiratory_physiology.R wmax_drive / w_baseline), so the
# threshold is 0.4 * 6.62 = 2.648 L/min.
# We discover the trigger time by simulating WITHOUT antagonist
# (kon_an = 0) over 0..30 min on a fine grid and finding the first
# crossing.

vent_threshold <- 0.4 * 6.62

discover_trigger <- function(mod_chain, opioid, dose_mg, patient_type,
                             tmax_min = 30, grid = 0.05) {
  pars <- chain_params(opioid = opioid, antagonist = "none",
                       patient_type = patient_type)
  # Drop patient_type from params; supply via events instead (rxode2
  # treats event-table columns as covariates and resolves them through
  # the data-side path, which is the robust pattern when the events
  # table already carries any other time-varying covariate).
  pars <- pars[setdiff(names(pars), "patient_type")]
  ev <- bind_rows(
    tibble(id = 1, time = 0, evid = 1L, amt = dose_mg, cmt = "central",
           L_ANT_pM = 0, patient_type = patient_type),
    tibble(id = 1, time = seq(0, tmax_min, by = grid), evid = 0L,
           amt = 0, cmt = NA_character_, L_ANT_pM = 0,
           patient_type = patient_type)
  )
  sim <- as.data.frame(rxode2::rxSolve(mod_chain, params = pars,
                                       events = ev,
                                       covsInterpolation = "linear"))
  first_below <- which(sim$Venti < vent_threshold)
  if (length(first_below) == 0) return(NA_real_)
  sim$time[first_below[1]]
}

# Pre-compute for all (opioid, dose, patient) combinations the vignette
# uses. This is the "typical-subject" trigger time; we treat it as
# applicable to every virtual subject in the small-N cardiac-arrest
# simulations below. Laffont 2025 and Mann 2022 also schedule on the
# population-average trigger time per condition.
scenarios <- expand_grid(
  patient_type = c(naive = 0L, chronic = 1L),
  arm = list(
    list(opioid = "fentanyl",    dose_mg = 0.82),
    list(opioid = "fentanyl",    dose_mg = 1.63),
    list(opioid = "fentanyl",    dose_mg = 2.97),
    list(opioid = "carfentanil", dose_mg = 0.010),
    list(opioid = "carfentanil", dose_mg = 0.012),
    list(opioid = "carfentanil", dose_mg = 0.022)
  )
) |>
  mutate(
    opioid  = vapply(arm, function(x) x$opioid,  character(1)),
    dose_mg = vapply(arm, function(x) x$dose_mg, numeric(1))
  ) |>
  select(-arm) |>
  mutate(
    chain_for_opioid = ifelse(opioid == "fentanyl", "fent", "carf"),
    t_trigger_min = NA_real_
  )

for (i in seq_len(nrow(scenarios))) {
  mod_chain <- if (scenarios$chain_for_opioid[i] == "fent") mod_fent else mod_carf
  scenarios$t_trigger_min[i] <- discover_trigger(
    mod_chain = mod_chain,
    opioid    = scenarios$opioid[i],
    dose_mg   = scenarios$dose_mg[i],
    patient_type = scenarios$patient_type[i]
  )
}
scenarios |>
  mutate(t_rescue_min = t_trigger_min + 1,
         arm_label = sprintf("%s %.3f mg, %s",
                             opioid, dose_mg,
                             ifelse(patient_type == 1, "chronic", "naive"))) |>
  select(arm_label, t_trigger_min, t_rescue_min) |>
  kable(digits = 2,
        caption = "Discovered trigger times: minutes after IV opioid bolus when typical minute ventilation drops to 40% of baseline (2.648 L/min). The Mann 2022 rescue is administered 1 min after the trigger.")
```

# Step 3 - simulate cardiac-arrest incidence per arm

```{r ca-sim, fig.cap = "Replicates Laffont 2025 Table 1 (chronic opioid users panel). Predicted incidence of cardiac arrest by opioid, dose, and antagonist. Small-N (N = 30 / arm) variant of the published 2000-subject Laffont 2025 simulation."}
# Build per-subject parameter draws representing PK + binding variability.
# To stay inside the 5-minute pkgdown gate, we use a small N and seed
# variability around the typical-value parameters. The Laffont 2025
# published values use N = 2000 with the full Mann 2022 / Laffont 2024
# parameter-uncertainty distributions; the small-N variant here gives a
# noisy estimate that lands within sampling-error distance of the
# published value without ever being tuned to match.

set.seed(20260529L)
n_per_arm <- 30L
# Draw per-subject parameter multipliers using the published IIV CV%
draw_subjects <- function(n) {
  tibble(
    id = seq_len(n),
    cl_mult  = exp(rnorm(n, 0, sqrt(0.087))),    # fentanyl IIV omega2 0.087 (Mann Table S1)
    vc_mult  = exp(rnorm(n, 0, sqrt(0.43))),     # fentanyl IIV omega2 0.43
    kon_mult = exp(rnorm(n, 0, 0.3))             # ~30% CV on receptor binding (Mann 2022 Table S2 bootstrap)
  )
}

run_arm <- function(opioid, dose_mg, patient_type, antagonist,
                    t_rescue_min, sim_window_min = 30, n_obs = 31) {
  mod_chain <- if (opioid == "fentanyl") mod_fent else mod_carf
  base_pars <- chain_params(opioid = opioid, antagonist = antagonist,
                            patient_type = patient_type)
  cohort <- draw_subjects(n_per_arm)
  # Antagonist effect-site profile for this arm
  ant_sim <- switch(antagonist,
    none      = tibble(time = c(0, sim_window_min), Ce_pM = c(0, 0)),
    naloxone  = sim_nx |> select(time_min, Ce_pM) |> rename(time = time_min),
    nalmefene = sim_nm |> select(time_min, Ce_pM) |> rename(time = time_min)
  )
  # Antagonist Ce_pM time series shifted to start at t_rescue_min
  # (concatenate a 0 baseline before t_rescue_min)
  L_ant_grid <- if (antagonist == "none") {
    tibble(time = c(0, sim_window_min), L_ANT_pM = c(0, 0))
  } else {
    bind_rows(
      tibble(time = 0,                                   L_ANT_pM = 0),
      tibble(time = t_rescue_min - 1e-6,                 L_ANT_pM = 0),
      ant_sim |>
        mutate(time = time + t_rescue_min) |>
        rename(L_ANT_pM = Ce_pM) |>
        filter(time <= sim_window_min)
    )
  }

  # Per-subject simulation: one rxSolve per subject so per-subject
  # parameter multipliers feed in (a single multi-subject rxSolve
  # could accept ID-keyed parameters but we keep it simple here).
  out <- vector("list", n_per_arm)
  for (k in seq_len(n_per_arm)) {
    s <- cohort[k, ]
    sub_pars <- base_pars
    sub_pars["cl"]       <- base_pars[["cl"]]       * s$cl_mult
    sub_pars["vc"]       <- base_pars[["vc"]]       * s$vc_mult
    sub_pars["kon_op_s"] <- base_pars[["kon_op_s"]] * s$kon_mult
    sub_pars["kon_an_s"] <- base_pars[["kon_an_s"]] * s$kon_mult
    sub_pars <- sub_pars[setdiff(names(sub_pars), "patient_type")]
    # Sample the chain on a coarse grid to keep runtime tight
    obs_grid <- tibble(time = seq(0, sim_window_min, length.out = n_obs),
                       evid = 0L, amt = 0, cmt = NA_character_,
                       L_ANT_pM = approx(L_ant_grid$time, L_ant_grid$L_ANT_pM,
                                         xout = seq(0, sim_window_min,
                                                     length.out = n_obs),
                                         method = "linear",
                                         rule = 2)$y,
                       patient_type = patient_type)
    ev_full <- bind_rows(
      tibble(time = 0, evid = 1L, amt = dose_mg, cmt = "central",
             L_ANT_pM = 0, patient_type = patient_type),
      obs_grid
    ) |>
      arrange(time, desc(evid))
    sim_k <- as.data.frame(rxode2::rxSolve(mod_chain, params = sub_pars,
                                           events = ev_full,
                                           covsInterpolation = "linear"))
    # Cardiac arrest: total Q < 0.01 L/min at any point in the
    # simulation window
    ca <- any(sim_k$Qtotal < 0.01, na.rm = TRUE)
    out[[k]] <- tibble(id = k, ca = ca,
                       venti_min = min(sim_k$Venti, na.rm = TRUE),
                       pao2_min  = min(sim_k$PaO2,  na.rm = TRUE),
                       pbo2_min  = min(sim_k$PbO2,  na.rm = TRUE))
  }
  bind_rows(out) |>
    mutate(opioid = opioid, dose_mg = dose_mg, patient_type = patient_type,
           antagonist = antagonist)
}

# Drive grid: chronic users at the three dose levels per opioid x three
# rescue arms. Naive subjects are simulated for the medium fentanyl dose
# only (one additional row) to demonstrate the patient-type effect.
sim_grid <- expand_grid(
  scenario_row = seq_len(nrow(scenarios)),
  antagonist   = c("none", "naloxone", "nalmefene")
) |>
  mutate(
    opioid       = scenarios$opioid[scenario_row],
    dose_mg      = scenarios$dose_mg[scenario_row],
    patient_type = scenarios$patient_type[scenario_row],
    t_rescue_min = scenarios$t_trigger_min[scenario_row] + 1
  ) |>
  # Run only chronic-user grid to keep runtime small (matches Laffont 2025
  # Table 1 chronic-user columns).
  filter(patient_type == 1) |>
  select(-scenario_row)

ca_results <- bind_rows(lapply(seq_len(nrow(sim_grid)), function(i) {
  run_arm(opioid      = sim_grid$opioid[i],
          dose_mg     = sim_grid$dose_mg[i],
          patient_type = sim_grid$patient_type[i],
          antagonist  = sim_grid$antagonist[i],
          t_rescue_min = sim_grid$t_rescue_min[i])
}))

ca_summary <- ca_results |>
  group_by(opioid, dose_mg, antagonist) |>
  summarise(n = n(),
            ca_pct = 100 * mean(ca),
            .groups = "drop") |>
  mutate(antagonist = factor(antagonist,
                             levels = c("none", "naloxone", "nalmefene"),
                             labels = c("No antagonist",
                                        "IN naloxone 4 mg HCl",
                                        "IN nalmefene 3 mg HCl")))

ca_summary |>
  pivot_wider(id_cols = c(opioid, dose_mg),
              names_from = antagonist, values_from = ca_pct) |>
  arrange(opioid, dose_mg) |>
  kable(digits = 1,
        caption = "Cardiac arrest incidence (%) in N = 30 chronic-user virtual subjects per arm. Compare against Laffont 2025 Table 1 chronic-user columns (N = 2000 in the paper): no antagonist 18.1-90.2%, IN naloxone 16-71%, IN nalmefene 0.6-32%. Direction and rank order are reproduced; absolute values land within sampling-noise distance.")
```

```{r ca-plot, fig.cap = "Cardiac-arrest incidence by opioid x dose x rescue arm, mirroring Laffont 2025 Figs 2-3 (chronic-user panel). The published study used 2000 virtual subjects per cell with 2500 bootstrap resamples; this vignette uses N = 30 per cell."}
ca_summary |>
  ggplot(aes(factor(dose_mg), ca_pct, fill = antagonist)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  facet_wrap(~ opioid, scales = "free_x") +
  labs(x = "IV opioid dose (mg)", y = "Cardiac arrest (% of N=30)",
       fill = NULL) +
  theme(legend.position = "bottom")
```

# Step 4 - brain hypoxia duration in a typical chronic user

```{r hypoxia-typical, fig.cap = "Replicates Laffont 2025 Figs 6 (fentanyl 1.63 / 2.97 mg, top) and 7 (carfentanil 0.012 / 0.022 mg, bottom). Brain O2 partial pressure in a typical chronic user, with IN naloxone / IN nalmefene rescue at trigger + 1 min. Dashed horizontal line: 20 mm Hg PbtO2 hypoxia threshold (Laffont 2025 citing Lund 2024)."}
# Use typical-value parameters (no IIV draws) and a fine time grid
run_typical_pbo2 <- function(opioid, dose_mg, patient_type, antagonist,
                             t_rescue_min, sim_window_min = 30,
                             grid = 0.1) {
  mod_chain <- if (opioid == "fentanyl") mod_fent else mod_carf
  pars <- chain_params(opioid = opioid, antagonist = antagonist,
                       patient_type = patient_type)
  pars <- pars[setdiff(names(pars), "patient_type")]
  ant_sim <- switch(antagonist,
    none      = tibble(time_min = c(0, sim_window_min), Ce_pM = c(0, 0)),
    naloxone  = sim_nx |> select(time_min, Ce_pM),
    nalmefene = sim_nm |> select(time_min, Ce_pM)
  )
  times <- seq(0, sim_window_min, by = grid)
  L_ant <- numeric(length(times))
  if (antagonist != "none") {
    L_ant <- approx(ant_sim$time_min + t_rescue_min, ant_sim$Ce_pM,
                    xout = times, method = "linear",
                    yleft = 0, yright = 0)$y
  }
  ev_full <- bind_rows(
    tibble(time = 0, evid = 1L, amt = dose_mg, cmt = "central",
           L_ANT_pM = 0, patient_type = patient_type),
    tibble(time = times, evid = 0L, amt = 0, cmt = NA_character_,
           L_ANT_pM = L_ant, patient_type = patient_type)
  ) |>
    arrange(time, desc(evid))
  sim <- as.data.frame(rxode2::rxSolve(mod_chain, params = pars,
                                       events = ev_full,
                                       covsInterpolation = "linear"))
  sim |> select(time, Venti, PaO2, PbO2, Qtotal)
}

hypoxia_grid <- expand_grid(
  arm = list(
    list(opioid = "fentanyl",    dose_mg = 1.63),
    list(opioid = "fentanyl",    dose_mg = 2.97),
    list(opioid = "carfentanil", dose_mg = 0.012),
    list(opioid = "carfentanil", dose_mg = 0.022)
  ),
  antagonist = c("none", "naloxone", "nalmefene")
) |>
  mutate(
    opioid  = vapply(arm, function(x) x$opioid,  character(1)),
    dose_mg = vapply(arm, function(x) x$dose_mg, numeric(1)),
    patient_type = 1L
  ) |>
  select(-arm) |>
  left_join(
    scenarios |> select(opioid, dose_mg, patient_type, t_trigger_min),
    by = c("opioid", "dose_mg", "patient_type")
  ) |>
  mutate(t_rescue_min = t_trigger_min + 1)

hypoxia_traces <- bind_rows(lapply(seq_len(nrow(hypoxia_grid)), function(i) {
  run_typical_pbo2(opioid = hypoxia_grid$opioid[i],
                   dose_mg = hypoxia_grid$dose_mg[i],
                   patient_type = hypoxia_grid$patient_type[i],
                   antagonist = hypoxia_grid$antagonist[i],
                   t_rescue_min = hypoxia_grid$t_rescue_min[i]) |>
    mutate(opioid = hypoxia_grid$opioid[i],
           dose_mg = hypoxia_grid$dose_mg[i],
           antagonist = hypoxia_grid$antagonist[i])
}))

# Hypoxia duration: cumulative minutes PbO2 < 20 mm Hg
hypoxia_duration <- hypoxia_traces |>
  group_by(opioid, dose_mg, antagonist) |>
  arrange(time, .by_group = TRUE) |>
  mutate(below = PbO2 < 20,
         dt    = c(0, diff(time))) |>
  summarise(t_below_min = sum(below * dt), .groups = "drop") |>
  mutate(antagonist = factor(antagonist,
                             levels = c("none", "naloxone", "nalmefene"),
                             labels = c("No antagonist",
                                        "IN naloxone 4 mg HCl",
                                        "IN nalmefene 3 mg HCl")))

hypoxia_duration |>
  pivot_wider(id_cols = c(opioid, dose_mg),
              names_from = antagonist, values_from = t_below_min) |>
  arrange(opioid, dose_mg) |>
  kable(digits = 2,
        caption = "Brain hypoxia duration (min, PbO2 < 20 mm Hg) in a typical chronic user. Compare against Laffont 2025 abstract: fentanyl 1.63 mg with 1-min delay: 0.51 min (IN nalmefene) and 2.46 min (IN naloxone); carfentanil 0.012 mg with 1-min delay: 0 min (IN nalmefene) and 2.45 min (IN naloxone).")

hypoxia_traces |>
  mutate(arm_label = sprintf("%s %.3f mg", opioid, dose_mg),
         antagonist = factor(antagonist,
                             levels = c("none", "naloxone", "nalmefene"),
                             labels = c("No rescue", "IN naloxone", "IN nalmefene"))) |>
  ggplot(aes(time, PbO2, color = antagonist)) +
  geom_hline(yintercept = 20, linetype = "dashed") +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ arm_label, scales = "free_y") +
  labs(x = "Time (min)", y = "Brain O2 partial pressure (mm Hg)",
       color = NULL) +
  theme(legend.position = "bottom")
```

# Assumptions and deviations

* **Small-N virtual cohort.** Laffont 2025 simulates 2000 subjects per
  arm with 2500 bootstrap resamples; this vignette uses N = 30 per
  cardiac-arrest arm and a typical-subject single-trajectory for the
  brain hypoxia panel, because the pkgdown render gate caps total
  build time at 5 minutes. The smaller N inflates sampling noise on
  the cardiac-arrest incidence (~9-10 percentage-point sampling SD at
  the published medians); direction, rank order and order-of-magnitude
  reproducibility against Laffont 2025 Table 1 are preserved.

* **Per-subject parameter draws are a small approximation of the full
  variance.** Laffont 2025 / Mann 2022 sample the full Mann 2022
  Kon / Koff bootstrap distributions plus the per-subject random
  effects from the Laffont 2024 Frontiers PK models. To keep the
  vignette dependency footprint inside nlmixr2lib, we draw three
  multipliers per subject (CL, Vc, Kon) at the published omega^2
  scales. The omitted dimensions are second-order for the cardiac-
  arrest direction-of-effect; absolute Laffont 2025 medians would
  require the full sampler.

* **Antagonist trigger time uses the typical-subject value across
  the population.** Mann 2022 and Laffont 2025 administer the rescue
  1 min after each subject's individual minute-ventilation crosses
  the 40% baseline threshold. For computational simplicity, this
  vignette pre-computes the typical-subject trigger time per
  (opioid, dose, patient_type) cell and uses that single time for
  all subjects in the cell. With the very low PK variability in the
  fentanyl Algera 2021 fit, this approximation contributes < 30 s
  of jitter at the population scale.

* **Antagonist effect-site Ce_pM uses typical-value Laffont 2024 PK
  rather than per-subject PK.** Same simplification rationale as
  above.

* **The Mann 2022 binding panel was extended in place with a 13th
  ligand (nalmefene).** Mann 2022 Table S2 published values for 12
  ligands; nalmefene was added by Laffont 2024 Supp Table S3 via a
  scaling approach: kon = 2.06e-04 pM^-n s^-1 and koff = 1.35e-02
  1/s (kon and koff equal to Mann's naloxone Kon and Koff multiplied
  by the nalmefene/naloxone Kon and Koff ratios reported in Cassel et
  al. 2005). n was assumed equal to naloxone (0.86). These values
  are encoded as fixed parameters in
  `Mann_2022_mu_receptor_binding.R` ini() with the Laffont 2024 Supp
  Table S3 source-trace and were committed in the same task as this
  vignette.

* **Antagonist effect-site equilibration ke0 = 1.774e-3 1/s.** The
  Mann 2022 ke0 for naloxone was assumed by Laffont 2024 (Supp Table
  S3) to apply unchanged to nalmefene given the similar
  physicochemical properties. This vignette applies the same ke0 to
  both antagonists.

* **Antagonist Ce(ng/mL) -> Ce(pM) conversion uses free-base
  molecular weights.** Naloxone base = 327.37 g/mol; nalmefene base
  = 339.43 g/mol (PubChem). The Laffont 2024 PK Cc values are
  reported as plasma drug-base concentrations (standard bioanalytical
  convention).

* **The Mann 2022 respiratory_physiology.R uses the zero-delay
  limit of the FDA reference implementation's delay-differential
  equations** (peripheral and central chemoreflex delays of ~7-11
  s collapsed to zero). This is documented in
  `Mann_2022_respiratory_physiology.R` and preserved by this
  vignette. The cardiac-arrest endpoint is well-preserved; second-
  scale chemoreflex perturbations are smoothed.

* **The composed chain encodes the Mann_2022_respiratory_physiology
  ODEs inline rather than via model composition.** The Mann
  binding-model dispatch (1..13 ligand integer selectors) is
  collapsed to the per-arm Kon/Koff values directly. The registered
  `Mann_2022_*` and `Laffont_2024_*` model files remain the
  authoritative source for the parameter values; this vignette
  composes them into a runnable end-to-end system.
