---
title: "Lemborexant (Lalovic 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Lemborexant (Lalovic 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Lalovic_2020_lemborexant"))$reference`
* Description: `r rxode2::rxode(readModelDb("Lalovic_2020_lemborexant"))$description`
* Article (open access): <https://doi.org/10.1002/jcph.1683>
* Supplemental information (Supplemental Tables S1-S2, Figures S1-S4):
  <https://doi.org/10.1002/jcph.1683>

Lemborexant is a dual orexin receptor antagonist approved for the treatment of
insomnia. Lalovic 2020 reports two analyses, and this package carries both:

* a **population PK model** built on 12 230 concentrations from 1892 subjects in
  12 phase 1-3 studies -- `Lalovic_2020_lemborexant`; and
* **exposure-response models** for the six most frequent treatment-emergent
  adverse events (TEAEs) in studies 202, 303 (SUNRISE 2) and 304 (SUNRISE 1),
  each a logistic regression on the individual average steady-state concentration
  (Cav,ss) -- `Lalovic_2020_lemborexant_somnolence`, `_nasopharyngitis`, `_uti`
  (urinary tract infection), `_influenza`, `_urti` (upper respiratory tract
  infection) and `_headache`.

The six adverse-event regressions were fitted independently, so each is its own
model file. The paper prints no regression coefficients for them: it tabulates
the fitted probabilities for a single reference subject (a 75-year-old White
woman) in Table 4. The coefficients here were back-solved from that table (see
the section on the exposure-response models below).

# Population

The PK analysis pooled 6 extensively and 3 sparsely sampled phase 1 studies, 1
phase 2 study and 2 phase 3 studies (Supplemental Table S1). Subjects were 18-88
years old (median 57), weighed 37-168 kg (median 74.1) with a BMI of 14.4-62.1
kg/m^2 (median 26.5), were 66% female and 70% White, and 547 of the 1892 were
elderly (Table 1). The phase 1 subjects were healthy adults and healthy elderly
subjects; the phase 2 and 3 subjects had insomnia disorder, and study 202
enrolled subjects with irregular sleep-wake rhythm disorder. Early studies used a
capsule, later ones the tablet (1755 of 1892 subjects). Doses spanned 1-100 mg; the
phase 3 doses were 5 and 10 mg at bedtime.

The exposure-response data set held 1664 subjects from studies 202 (n = 62), 303
(n = 726) and 304 (n = 524), on placebo or 5-15 mg lemborexant.

```{r population}
str(rxode2::rxode(readModelDb("Lalovic_2020_lemborexant"))$population)
```

# Source trace

| Item | Value | Source location |
|---|---|---|
| Structure: 3 compartments, linear elimination | depot, central, 2 peripherals | Results, Base Model; Supplemental Figure S2 |
| Absorption: lag, then zero-order (D1) into depot, then first-order (ka) | ADVAN12, D1 on CMT 1 | Results, Base Model; Supplemental Figure S2 |
| `lcl` CL/F | 22.7 L/h | Table 2 |
| `lvc` V2/F | 9.09 L | Table 2 |
| `lq` Q3/F | 32.1 L/h | Table 2 |
| `lvp` V3/F | 278 L | Table 2 |
| `lq2` Q4/F | 31.0 L/h | Table 2 |
| `lvp2` V4/F | 783 L | Table 2 |
| `ld1` D1, capsule | 0.467 h (fixed) | Table 2 |
| `lka` Ka, capsule | 0.532 1/h (fixed) | Table 2 |
| `ltlag` ALAG1 | 0.403 h (fixed) | Table 2 |
| `lfdepot` F1 | 1 (fasted reference) | Table 2 lists only the food effect |
| `e_bmi_cl` | -0.428, (BMI/25)^theta | Table 2; reference back-solved (see Assumptions) |
| `e_alp_cl` | -0.118, (ALP/71)^theta | Table 2; reference = Table 1 median and Table 3 footnote |
| `e_age_gt65_cl` | 0.739^elderly | Table 2 |
| `e_form_tablet_d1` | 0.254 (fixed) | Table 2; Results give 0.118 h for the tablet |
| `e_dosetime_evening_d1` | 2.33 (fixed) | Table 2 'D1, nighttime dosing' |
| `e_form_tablet_ka` | 1.12 (fixed) | Table 2 |
| `e_fed_highfat_ka` | 0.695 (fixed) | Table 2 |
| `e_fed_highfat_fdepot` | 1.21 (fixed) | Table 2 |
| IIV, CL/F V2/F Q3/F V3/F Q4/F V4/F | 48.1, 142, 56.8, 82.0, 46.5, 41.4 %CV | Table 2; variance = (CV/100)^2 per the Table 2 abbreviation 'CV, square root of variance x 100' |
| IIV, D1 Ka F1 (fixed) | 167, 43.8, 68.1 %CV | Table 2 |
| `propSd_early`, `addSd_early` (TAD <= 3 h) | 0.329, 2.62 ng/mL | Table 2 |
| `propSd_late`, `addSd_late` (TAD > 3 h) | 0.143, 0.0189 ng/mL | Table 2 |
| Covariate forms: power for continuous, theta^indicator for dichotomous | -- | Table 2 footnote |
| TEAE models: `logit_placebo` | qlogis(placebo row) | Table 4, Placebo row |
| TEAE models: `logit_active`, `e_cav_logit` | back-solved | Table 4, ten 5 mg and 10 mg rows (see below) |

# The population PK model

## Typical-value profiles

The phase 3 regimen is the tablet taken at bedtime, fasted. For the reference
adult (BMI 25 kg/m^2, ALP 71 U/L) that gives D1 = 0.467 x 0.254 x 2.33 = 0.276 h
and ka = 0.532 x 1.12 = 0.596 1/h. Every dose record carries `rate = -2`, which is
what makes rxode2 honour the modelled duration `dur(depot)`.

```{r typical-setup}
mod <- readModelDb("Lalovic_2020_lemborexant")
mod_typ <- rxode2::zeroRe(mod)

ref_cov <- data.frame(
  BMI = 25, ALP = 71, AGE_GT65 = 0,
  FORM_TABLET = 1, FED_HIGHFAT = 0, DOSETIME_EVENING = 1
)

# Once-daily dosing for n_days days, observations over one chosen interval.
ss_events <- function(dose, n_days, obs_day, cov, step = 0.05) {
  doses <- data.frame(
    time = 24 * (seq_len(n_days) - 1), amt = dose, evid = 1L,
    rate = -2, cmt = "depot"
  )
  obs <- data.frame(
    time = 24 * (obs_day - 1) + seq(0, 24, by = step), amt = 0, evid = 0L,
    rate = 0, cmt = "central"
  )
  ev <- dplyr::bind_rows(doses, obs) |> dplyr::arrange(time, dplyr::desc(evid))
  ev$id <- 1L
  cbind(ev, cov[rep(1, nrow(ev)), , drop = FALSE])
}

trap <- function(t, y) sum(diff(t) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)

n_days <- 42
s_d1 <- rxode2::rxSolve(mod_typ, ss_events(5, 1, 1, ref_cov), returnType = "data.frame")
s_ss <- rxode2::rxSolve(mod_typ, ss_events(5, n_days, n_days, ref_cov), returnType = "data.frame")
s_prev <- rxode2::rxSolve(mod_typ, ss_events(5, n_days, n_days - 1, ref_cov), returnType = "data.frame")
s_d1 <- s_d1[!is.na(s_d1$Cc), ]
s_ss <- s_ss[!is.na(s_ss$Cc), ]
s_prev <- s_prev[!is.na(s_prev$Cc), ]
t_ss <- s_ss$time - min(s_ss$time)

auc_d1 <- trap(s_d1$time, s_d1$Cc)
auc_ss <- trap(s_ss$time, s_ss$Cc)
auc_prev <- trap(s_prev$time, s_prev$Cc)
cav_ss <- auc_ss / 24
```

```{r fig-typical, fig.cap = "Typical-value lemborexant concentrations after the first 5 mg bedtime tablet dose and at steady state, reference adult (BMI 25 kg/m^2, ALP 71 U/L)."}
dplyr::bind_rows(
  data.frame(t = s_d1$time, Cc = s_d1$Cc, Profile = "Day 1"),
  data.frame(t = t_ss, Cc = s_ss$Cc, Profile = "Steady state")
) |>
  ggplot2::ggplot(ggplot2::aes(t, Cc, colour = Profile)) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::labs(x = "Time after dose (h)", y = "Lemborexant (ng/mL)", colour = NULL) +
  ggplot2::theme_bw()
```

Three checks follow. The first is an identity: at steady state the average
concentration must equal F x dose / (24 x CL/F), a statement about the model
itself that does not depend on the paper. The other two compare with numbers
stated in the Discussion of Lalovic 2020.

```{r typical-checks}
cav_closed <- 1000 * 5 / (24 * 22.7) # ng/mL; F = 1 and CL/F = 22.7 L/h at the reference
c_at <- function(h) stats::approx(t_ss, s_ss$Cc, xout = h)$y
cum_auc <- function(h) trap(t_ss[t_ss <= h], s_ss$Cc[t_ss <= h])

typ_tab <- data.frame(
  Quantity = c(
    "Cav,ss (ng/mL), simulated",
    "Cav,ss (ng/mL), dose / (24 x CL/F)",
    "Accumulation ratio AUCtau,ss / AUC0-24 day 1",
    "Fraction of AUCtau,ss eliminated by 8 h",
    "Fraction of AUCtau,ss eliminated by 9 h",
    "Cmax,ss / C(8 h)",
    "Cmax,ss / C(9 h)"
  ),
  Model = c(
    cav_ss, cav_closed, auc_ss / auc_d1, cum_auc(8) / auc_ss,
    cum_auc(9) / auc_ss, max(s_ss$Cc) / c_at(8), max(s_ss$Cc) / c_at(9)
  ),
  Paper = c("--", "--", "1.9", "0.53-0.57", "0.53-0.57", "3.5-4", "3.5-4")
)
typ_tab$Model <- signif(typ_tab$Model, 3)
knitr::kable(typ_tab, caption = "Typical-value checks against the Discussion of Lalovic 2020.")

stopifnot(
  # Steady state reached: successive intervals agree.
  abs(auc_ss / auc_prev - 1) < 1e-3,
  # Identity: steady-state average concentration = dose / (tau x CL/F).
  abs(cav_ss / cav_closed - 1) < 0.005,
  # Discussion: accumulation ratio 1.9 on once-daily bedtime dosing.
  abs(auc_ss / auc_d1 - 1.9) < 0.2,
  # Discussion: 53%-57% of AUCss eliminated 8-9 h after the dose.
  cum_auc(8) / auc_ss > 0.50, cum_auc(9) / auc_ss < 0.60
)
```

The model reproduces the accumulation ratio and the fraction of the steady-state
AUC gone by the morning. The peak-to-morning ratio comes out at about 4.3-4.5 for
the typical subject against the 3.5- to 4-fold quoted "on average"; the paper's
figure is a mean over individuals, and the very large between-subject variability
on V2/F (142% CV) and D1 (167% CV) flattens individual peaks, so the typical value
sits above the average ratio. It is not used as a gate.

## Covariate effects (replicates the Results text and Table 3)

The Results quote the size of each clearance covariate relative to the typical
subject. They are closed-form consequences of Table 2 once the reference values
are fixed, so they check the transcription and the unprinted BMI reference (25
kg/m^2; see Assumptions).

```{r cov-effects}
cl_rel <- function(BMI = 25, ALP = 71, AGE_GT65 = 0) {
  (BMI / 25)^-0.428 * (ALP / 71)^-0.118 * 0.739^AGE_GT65
}
eff <- data.frame(
  Statement = c(
    "Elderly: CL/F lower by", "ALP 150 U/L: exposure higher by",
    "ALP 35 U/L: exposure lower by", "BMI 32: exposure higher by",
    "BMI 40: exposure higher by", "BMI 15: CL/F higher by"
  ),
  Model_pct = 100 * c(
    1 - cl_rel(AGE_GT65 = 1), 1 / cl_rel(ALP = 150) - 1,
    1 - 1 / cl_rel(ALP = 35), 1 / cl_rel(BMI = 32) - 1,
    1 / cl_rel(BMI = 40) - 1, cl_rel(BMI = 15) - 1
  ),
  Paper_pct = c(26, 9, 9, 11, 22, 25)
)
eff$Model_pct <- round(eff$Model_pct, 1)
eff |>
  dplyr::rename(`Results statement` = Statement, `Model (%)` = Model_pct, `Paper (%)` = Paper_pct) |>
  knitr::kable(caption = "Covariate effect sizes quoted in the Results ('Clinically Relevant Covariates').")
stopifnot(all(abs(eff$Model_pct - eff$Paper_pct) < 1.5))
```

Table 3 compares steady-state AUC across BMI classes and for overweight elderly
versus normal-weight adults at 5 mg nightly, with ALP at its 71 U/L median. The
published ratios come from 250 resampled subjects, so the typical-value ratios
below are not expected to match to the last percent.

```{r table3}
profiles <- data.frame(
  Profile = c("Normal (BMI 22.9)", "Underweight (BMI 17.4)", "Overweight (BMI 27.4)",
              "Obese (BMI 32.7)", "Elderly, overweight (BMI 26.6)"),
  BMI = c(22.9, 17.4, 27.4, 32.7, 26.6),
  AGE_GT65 = c(0, 0, 0, 0, 1),
  Paper = c(NA, 89, 111, 119, 139)
)
profiles$AUCss <- vapply(seq_len(nrow(profiles)), function(i) {
  cv <- ref_cov
  cv$BMI <- profiles$BMI[i]
  cv$AGE_GT65 <- profiles$AGE_GT65[i]
  s <- rxode2::rxSolve(mod_typ, ss_events(5, n_days, n_days, cv), returnType = "data.frame")
  s <- s[!is.na(s$Cc), ]
  trap(s$time, s$Cc)
}, numeric(1))
profiles$Model <- round(100 * profiles$AUCss / profiles$AUCss[1], 1)
profiles |>
  dplyr::select(Profile, Model, Paper) |>
  dplyr::rename(`Model AUCss ratio (%)` = Model, `Table 3 ratio (%)` = Paper) |>
  knitr::kable(caption = "Replicates Table 3 of Lalovic 2020 (reference: normal-BMI adult).")
chk3 <- profiles[-1, ]
stopifnot(all(abs(chk3$Model / chk3$Paper - 1) < 0.05))
```

The underweight ratio matches exactly; the overweight and obese ratios fall 2-3
percentage points below the published ones and the elderly-overweight ratio 5
points above (144% against 139%), within 4% in relative terms and the scale
expected from resampling 250 subjects.

## Stochastic simulation of the phase 3 regimen

A virtual cohort of 200 subjects per arm takes 5 or 10 mg tablets at bedtime,
fasted, for 42 days. BMI and ALP are log-normal around the Table 1 medians
(26.5 kg/m^2, 71 U/L) with coefficients of variation of 18% and 30% (from the
Table 1 SDs), redrawn when outside the Table 1 ranges, and 29% of subjects are
elderly (547/1892, Table 1).

```{r cohort}
set.seed(20201201)
rxode2::rxSetSeed(20201201)
n_per_arm <- 200

draw_in_range <- function(n, median, cv, lo, hi) {
  x <- stats::rlnorm(n, log(median), sqrt(log(1 + cv^2)))
  bad <- x < lo | x > hi
  while (any(bad)) {
    x[bad] <- stats::rlnorm(sum(bad), log(median), sqrt(log(1 + cv^2)))
    bad <- x < lo | x > hi
  }
  x
}

make_cohort <- function(dose, id_offset) {
  n <- n_per_arm
  data.frame(
    id = id_offset + seq_len(n),
    dose = dose,
    treatment = paste(dose, "mg"),
    BMI = draw_in_range(n, 26.5, 0.185, 14.4, 62.1),
    ALP = draw_in_range(n, 71, 0.30, 13, 256),
    AGE_GT65 = stats::rbinom(n, 1, 547 / 1892),
    FORM_TABLET = 1, FED_HIGHFAT = 0, DOSETIME_EVENING = 1
  )
}
cohort <- dplyr::bind_rows(make_cohort(5, 0L), make_cohort(10, n_per_arm))

obs_start <- 24 * (n_days - 1)
dose_rows <- cohort |>
  dplyr::select(id, dose) |>
  tidyr::crossing(day = seq_len(n_days)) |>
  dplyr::transmute(id, time = 24 * (day - 1), amt = dose, evid = 1L, rate = -2, cmt = "depot")
obs_rows <- cohort |>
  dplyr::select(id) |>
  tidyr::crossing(time = obs_start + c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 9, 12, 16, 20, 24)) |>
  dplyr::mutate(amt = 0, evid = 0L, rate = 0, cmt = "central")
events <- dplyr::bind_rows(dose_rows, obs_rows) |>
  dplyr::left_join(cohort, by = "id") |>
  dplyr::arrange(id, time, dplyr::desc(evid))
stopifnot(!anyDuplicated(cohort$id))

sim <- rxode2::rxSolve(
  mod, events,
  keep = c("treatment", "dose"), returnType = "data.frame"
)
```

```{r fig-vpc, fig.cap = "Simulated steady-state lemborexant concentrations (median and 5th-95th percentiles) over the last 24 h dosing interval of 5 and 10 mg at bedtime. Lalovic 2020 Figure 2 shows the corresponding prediction-corrected VPCs stratified by study."}
sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(tad = time - obs_start) |>
  dplyr::group_by(treatment, tad) |>
  dplyr::summarise(
    p05 = stats::quantile(Cc, 0.05), p50 = stats::median(Cc), p95 = stats::quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot2::ggplot(ggplot2::aes(tad, p50, colour = treatment, fill = treatment)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = p05, ymax = p95), alpha = 0.2, colour = NA) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::scale_y_log10() +
  ggplot2::labs(x = "Time after dose (h)", y = "Lemborexant (ng/mL)", colour = NULL, fill = NULL) +
  ggplot2::theme_bw()
```

```{r fig-cl, fig.cap = "Individual CL/F by elderly status and BMI class in the virtual cohort; compare with Lalovic 2020 Figure 3."}
sim |>
  dplyr::distinct(id, .keep_all = TRUE) |>
  dplyr::left_join(dplyr::select(cohort, id, BMI_c = BMI, AGE_c = AGE_GT65), by = "id") |>
  dplyr::mutate(
    Age = ifelse(AGE_c == 1, "Elderly (> 65 y)", "Adult"),
    `BMI class` = cut(BMI_c, c(0, 18.5, 25, 30, Inf), right = FALSE,
                      labels = c("< 18.5", "18.5-24.9", "25-29.9", ">= 30"))
  ) |>
  ggplot2::ggplot(ggplot2::aes(`BMI class`, cl, fill = Age)) +
  ggplot2::geom_boxplot(outlier.size = 0.5) +
  ggplot2::scale_y_log10() +
  ggplot2::labs(y = "Individual CL/F (L/h)", fill = NULL) +
  ggplot2::theme_bw()
```

## PKNCA validation

The paper reports no NCA table, so PKNCA is used here to characterise the
steady-state interval and to check the steady-state identity AUCtau = F x dose /
(CL/F) subject by subject, with the individual F and CL/F from the solve.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id, concu = "ng/mL", timeu = "h")
dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id, doseu = "mg")

intervals <- data.frame(
  start = obs_start, end = obs_start + 24,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, cav = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_tab <- as.data.frame(nca_res)

nca_tab |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "cmin", "auclast", "cav")) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(median = signif(stats::median(PPORRES), 3), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  dplyr::rename(
    Treatment = treatment, `Cmax,ss (ng/mL)` = cmax, `Tmax (h)` = tmax,
    `Cmin,ss (ng/mL)` = cmin, `AUCtau,ss (ng*h/mL)` = auclast, `Cav,ss (ng/mL)` = cav
  ) |>
  knitr::kable(caption = "Median steady-state NCA over the last 24 h interval (simulated).")

ind <- sim |>
  dplyr::distinct(id, .keep_all = TRUE) |>
  dplyr::select(id, cl, fdepot, dose)
idchk <- nca_tab |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::left_join(ind, by = "id") |>
  dplyr::mutate(ratio = PPORRES / (1000 * fdepot * dose / cl))
stopifnot(
  nrow(idchk) == 2 * n_per_arm,
  # Sparse 16-point grid: trapezoid error on the narrow peak moves the ratio a
  # few percent per subject; a transcribed CL, F or unit error moves it by tens.
  abs(stats::median(idchk$ratio) - 1) < 0.05,
  stats::quantile(abs(idchk$ratio - 1), 0.9) < 0.15
)
```

The individual Cav,ss is what drives the exposure-response models. Table 4
of Lalovic 2020 prints its 5th/25th/50th/75th/95th percentiles in the phase 2-3
population: 6.3/7.9/10/14/19 ng/mL at 5 mg and 12/17/21/27/37 ng/mL at 10 mg.

```{r cav-quantiles}
cav_q <- idchk |>
  dplyr::mutate(cav = PPORRES / 24) |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    p05 = stats::quantile(cav, 0.05), p25 = stats::quantile(cav, 0.25),
    p50 = stats::median(cav), p75 = stats::quantile(cav, 0.75), p95 = stats::quantile(cav, 0.95)
  )
table4_q <- data.frame(
  treatment = c("5 mg", "10 mg"),
  t05 = c(6.3, 12), t25 = c(7.9, 17), t50 = c(10, 21), t75 = c(14, 27), t95 = c(19, 37)
)
cav_cmp <- dplyr::left_join(cav_q, table4_q, by = "treatment")
cav_cmp |>
  dplyr::mutate(dplyr::across(p05:p95, ~ signif(.x, 3))) |>
  dplyr::rename(
    Treatment = treatment, `Sim P5` = p05, `Sim P25` = p25, `Sim median` = p50,
    `Sim P75` = p75, `Sim P95` = p95, `Table 4 P5` = t05, `Table 4 P25` = t25,
    `Table 4 median` = t50, `Table 4 P75` = t75, `Table 4 P95` = t95
  ) |>
  knitr::kable(caption = "Cav,ss percentiles: virtual cohort versus Lalovic 2020 Table 4.")
stopifnot(all(abs(cav_cmp$p50 / cav_cmp$t50 - 1) < 0.2))
```

The medians agree within about 10%. The simulated tails are much wider than the
Table 4 ones. Table 4
was built from empirical Bayes estimates in sparsely sampled phase 3 subjects,
which shrink toward the typical value, while the simulation draws the full
between-subject variability, including the 68% CV on relative bioavailability.

# The exposure-response models

## How the coefficients were obtained

Each TEAE regression in Lalovic 2020 has the form logit(p) = intercept + slope x
Cav,ss + age, sex and race terms, fitted to the pooled placebo and active data.
None of the coefficients is printed. Table 4 instead prints the fitted probability
for a 75-year-old White woman at ten Cav,ss values (the 5th-95th percentiles at 5
and 10 mg) and on placebo. With age, sex and race fixed at those values the
logit is a straight line in Cav,ss, so the maintainers recovered the intercept
and slope for that subject by least squares on qlogis(probability) against the
printed Cav,ss over the ten active rows. The chunk below repeats that derivation
from the transcribed table and checks that it gives the values in the model files.

```{r table4}
table4 <- data.frame(
  cav = c(0, 6.3, 7.9, 10, 14, 19, 12, 17, 21, 27, 37),
  arm = c("Placebo", rep("5 mg", 5), rep("10 mg", 5)),
  somnolence = c(5.3, 7.2, 7.4, 7.7, 8.1, 8.9, 7.7, 8.4, 8.9, 9.6, 11.0),
  nasopharyngitis = c(8.6, 10.0, 9.5, 8.9, 8.1, 6.7, 8.9, 7.6, 6.7, 5.8, 4.0),
  uti = c(2.5, 2.0, 2.1, 2.3, 2.5, 2.9, 2.2, 2.6, 2.9, 3.3, 4.4),
  influenza = c(4.0, 3.8, 3.6, 3.4, 3.1, 2.6, 3.4, 2.9, 2.6, 2.3, 1.7),
  urti = c(3.2, 4.7, 4.4, 4.1, 3.6, 3.0, 4.1, 3.4, 3.0, 2.5, 1.7),
  headache = c(8.2, 11.0, 9.8, 9.2, 8.3, 6.8, 9.2, 7.7, 6.8, 5.7, 3.9)
)
endpoints <- c("somnolence", "nasopharyngitis", "uti", "influenza", "urti", "headache")
active <- table4$arm != "Placebo"

er_ui <- lapply(
  stats::setNames(endpoints, endpoints),
  function(ep) rxode2::rxode(readModelDb(paste0("Lalovic_2020_lemborexant_", ep)))
)

coef_chk <- dplyr::bind_rows(lapply(endpoints, function(ep) {
  y <- stats::qlogis(table4[[ep]] / 100)
  fit <- stats::lm(y[active] ~ table4$cav[active])
  th <- er_ui[[ep]]$theta
  data.frame(
    endpoint = ep,
    refit_intercept = unname(stats::coef(fit)[1]), model_intercept = th[["logit_active"]],
    refit_slope = unname(stats::coef(fit)[2]), model_slope = th[["e_cav_logit"]],
    table_placebo = y[!active], model_placebo = th[["logit_placebo"]]
  )
}))
knitr::kable(coef_chk, digits = 5, caption = "Back-solved coefficients: re-derived from Table 4 versus the values in the model files.")
stopifnot(
  # Same computation on both sides; the model files store 4-5 significant figures.
  all(abs(coef_chk$refit_intercept - coef_chk$model_intercept) < 5e-4),
  all(abs(coef_chk$refit_slope - coef_chk$model_slope) < 5e-5),
  all(abs(coef_chk$table_placebo - coef_chk$model_placebo) < 5e-4)
)
```

## Reproducing Table 4

```{r table4-repro}
solve_er <- function(ep, cav, placebo) {
  ev <- data.frame(id = seq_along(cav), time = 0, evid = 0L, CAV = cav, PLACEBO = placebo)
  s <- rxode2::rxSolve(er_ui[[ep]], ev, returnType = "data.frame")
  100 * s[[paste0("prob_", ep)]]
}
repro <- table4[, c("cav", "arm")]
for (ep in endpoints) {
  repro[[ep]] <- solve_er(ep, table4$cav, as.integer(!active))
}
err <- as.matrix(repro[, endpoints]) - as.matrix(table4[, endpoints])

repro_show <- repro
repro_show[, endpoints] <- round(repro_show[, endpoints], 1)
repro_show |>
  dplyr::rename(
    `Cav,ss (ng/mL)` = cav, Arm = arm, Somnolence = somnolence,
    Nasopharyngitis = nasopharyngitis, `Urinary tract infection` = uti,
    Influenza = influenza, `Upper respiratory tract infection` = urti, Headache = headache
  ) |>
  knitr::kable(caption = "Replicates Table 4 of Lalovic 2020: model-predicted TEAE probability (%) for a 75-year-old White woman.")

cat("Largest absolute difference from Table 4:", round(max(abs(err)), 2), "percentage points\n")
stopifnot(
  all(abs(err[!active, ]) < 0.01), # placebo row reproduced exactly
  max(abs(err[active, ])) < 0.5 # active rows within the rounding of the printed Cav,ss
)
```

All 66 cells are reproduced within 0.4 percentage points; the remaining
difference reflects the two-significant-figure Cav,ss column. The table itself is
not fully consistent at that precision: for every endpoint, the 10 mg 5th
percentile (printed Cav,ss 12 ng/mL) shows the same probability as the 5 mg median
(10 ng/mL), and the 10 mg median (21 ng/mL) the same as the 5 mg 95th percentile
(19 ng/mL).

```{r fig-er, fig.cap = "TEAE probability versus Cav,ss for a 75-year-old White woman (lines), with the Table 4 points (active: circles; placebo: triangles at Cav,ss = 0). The flat relationships are the paper's main exposure-response finding."}
grid <- seq(0, 40, by = 0.5)
curves <- dplyr::bind_rows(lapply(endpoints, function(ep) {
  data.frame(endpoint = ep, cav = grid, p = solve_er(ep, grid, 0L))
}))
pts <- table4 |>
  tidyr::pivot_longer(dplyr::all_of(endpoints), names_to = "endpoint", values_to = "p") |>
  dplyr::mutate(Arm = ifelse(arm == "Placebo", "Placebo", "Lemborexant"))
ggplot2::ggplot(curves, ggplot2::aes(cav, p)) +
  ggplot2::geom_line() +
  ggplot2::geom_point(data = pts, ggplot2::aes(shape = Arm)) +
  ggplot2::facet_wrap(~endpoint, scales = "free_y") +
  ggplot2::scale_shape_manual(values = c(Lemborexant = 1, Placebo = 17)) +
  ggplot2::labs(x = "Cav,ss (ng/mL)", y = "Probability (%)", shape = NULL) +
  ggplot2::theme_bw()
```

# Assumptions and deviations

- **BMI reference value (25 kg/m^2) is not printed.** Table 2 gives the BMI
  exponent (-0.428) and the footnote states a power form around a "generally
  centered reference value" without naming it. The maintainers back-solved the
  reference from the three BMI effect sizes in the Results (BMI 32: 11% and BMI 40:
  22% higher exposure; BMI 15: 25% change in CL/F relative to "the typical
  subject CL/F value with a normal BMI"). 25 kg/m^2 reproduces all three; the
  cohort median of 26.5 kg/m^2 does not (8%, 19%, 28%), nor do 24 or 26 kg/m^2.
  The Table 3 ratios are independent of this choice.
- **ALP reference value (71 U/L).** Not printed in Table 2; 71 U/L is the Table 1
  median and the value the authors assumed for the Table 3 simulations, and it
  reproduces the quoted ALP effect sizes.
- **Elderly cut point.** The Methods, the Results text and the Table 3 categories
  define elderly as > 65 years, which is what `AGE_GT65` encodes. Table 1 and the
  Discussion write >= 65 years; the difference affects only subjects aged
  exactly 65.
- **Final model versus base model.** The absorption parameters and their
  covariate effects are fixed in the final model (Table 2) "from the base model",
  but the values in Table 2 (e.g. D1 capsule 0.467 h, tablet factor 0.254,
  nighttime factor 2.33, Ka food factor 0.695, F1 food factor 1.21) differ from
  the base-model estimates in Supplemental Table S2 (0.598 h, 0.162, 1.20, 0.626,
  1.28), and Table S2 has no IIV on F1 while Table 2 does. The Results text quotes
  the Table 2 values (0.118 h tablet D1, 2.33-fold bedtime D1, 12% higher Ka on
  tablet, 30% lower Ka and 21% higher F1 with food), so Table 2 is used
  throughout. The base-model dose >= 50 mg factor on D1 (Table S2) was dropped
  from the final model (Results, Final Model) and is not included.
- **Sequential zero- then first-order absorption.** The paper describes "mixed
  first-order (Ka) and zero-order (D1) absorption with a lag time" in NONMEM ADVAN12,
  and Supplemental Figure S2 shows a single depot-to-central arrow labelled
  "ka/D1". No split fraction is reported, which identifies the usual ADVAN12 form
  with D1 on the depot: the dose enters the depot at a constant rate over D1 and
  is absorbed first-order from it. Dose records must carry `rate = -2`.
- **Relative bioavailability.** F1 is 1 in the fasted state (Table 2 reports only
  the food factor) and carries a fixed log-normal IIV of 68.1% CV.
- **IIV and residual error scale.** Table 2 reports %CV and defines CV as the
  square root of the variance x 100, so each variance is (CV/100)^2 and each
  residual proportional SD is CV/100.
- **Food covariate.** The Methods describe the food covariate as a standard
  high-fat meal, so it is carried as `FED_HIGHFAT`.
- **Bedtime dosing** is carried as `DOSETIME_EVENING` (the dose-record property of
  being taken at bedtime); the paper gives no clock-time window.
- **Covariates screened but not retained.** PPI use, sex, body weight and
  creatinine clearance are listed in `covariatesDataExcluded`; race, dose and the
  liver markers other than ALP were also screened and did not enter the final
  model (Results; Supplemental Figure S4).
- **Exposure-response coefficients are back-solved, not printed.** See the
  derivation above. Because Table 4 fixes age (75 years), sex (female) and race
  (White), the models represent that reference subject only; the source's age,
  sex and race coefficients are not reported and cannot be recovered. The slope is
  supported only over the active Cav,ss range in Table 4 (about 6-37 ng/mL).
- **Placebo in the exposure-response models.** The Table 4 placebo row does not
  lie on the straight logit line through the active rows at Cav,ss = 0, so it is
  carried as its own logit, selected by `PLACEBO = 1`. Its six values equal
  the pooled incidences the Results text labels "active treatment" (for example
  somnolence "8.5% placebo, 5.3% active"), which suggests that the text swapped the
  two labels and that the Table 4 placebo row is the observed placebo proportion.
  The Discussion (higher somnolence on lemborexant than on placebo in both SUNRISE
  studies) supports that reading.
- **Placeholder residual error on the probabilities.** The logistic regressions
  have a Bernoulli likelihood with no residual term; each file carries a fixed
  0.001 additive SD only so that rxode2 has an error model to attach.
- **No correction notice** for Lalovic 2020 was found on EuropePMC as of
  2026-09-27.
