---
title: "Ondansetron (Lam 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ondansetron (Lam 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
mod <- readModelDb("Lam_2025_ondansetron")
ui <- rxode2::rxode(mod)
```

* Citation: `r ui$reference`
* Article: <https://doi.org/10.1111/cts.70147>
* Supplement (Appendix S1, NONMEM control stream of the final model): <https://doi.org/10.1111/cts.70147>

Lam 2025 characterises ondansetron disposition in neonates with neonatal
opioid withdrawal syndrome (NOWS). Because the neonatal PK data are sparse
(109 plasma samples from 36 neonates, median 3 samples per neonate), the
authors did not fit a de novo model. Instead they updated a published
two-compartment intravenous ondansetron model in infants aged 1-48 months
(the "infant reference model") with Metropolis-Hastings MCMC in NONMEM
(`METHOD=BAYES`), supplying log-scale normal priors for CL, V, V2 and Q that
were computed from the reference model at the cohort median birth weight and
postmenstrual age, and a weakly informative prior for the previously
uncharacterised oral absorption rate constant KA.

The key finding is a much slower absorption rate in neonates
(KA = 0.19 /h, i.e. an apparent absorption half-life of about 3.6 h) than the
adult assumption used in the pre-trial simulations.

## What the final model does and does not include

The paper reports a model-selection exercise (Table 2) that compared a
covariate model (allometric scaling plus a clearance-maturation term applied
at the individual level) against a model with no individual-level covariate
effects, using PSIS-LOO cross-validation. The model **without** covariate
effects was selected: the covariate model underpredicted individual
concentrations while performing similarly on individual fits and individual
posterior predictive checks (Results 3.1). The Appendix S1 `$PK` block of the
final run confirms this - it is `CL=EXP(MU_1+ETA(1))`, `V=EXP(XMU2)`,
`V2=EXP(XMU3)`, `CLD=EXP(XMU4)`, `KA=EXP(XMU5)`, with no weight or age terms.

Allometric scaling and clearance maturation therefore enter the analysis
**only through the priors**, evaluated once at the cohort median birth weight
(3.19 kg) and median postmenstrual age (39 1/7 weeks).

The final model therefore has **no covariate effect on any structural
parameter**. It does read exactly one covariate, and not as a parameter
effect: the mother's plasma concentration at delivery, which sets the
neonate's central-compartment initial condition (`A_0(2) = C0 * S2`).

```{r covariates}
data.frame(
  Covariate = names(ui$meta$covariateData),
  Units = vapply(ui$meta$covariateData, function(x) x$units, character(1)),
  Role = "Initial condition on `central` at t = 0 (transplacental transfer)",
  `Source column` = vapply(ui$meta$covariateData, function(x) x$source_name,
                           character(1)),
  row.names = NULL, check.names = FALSE
) |>
  knitr::kable(caption = "The only covariate the Lam 2025 final model reads.")
```

Every demographic the paper screened but did not retain is recorded in the
model's `covariatesDataExcluded` metadata for provenance.

```{r excluded-covariates}
data.frame(
  Covariate = names(ui$meta$covariatesDataExcluded),
  Units = vapply(ui$meta$covariatesDataExcluded, function(x) x$units, character(1)),
  Role = c(
    "Prior only (allometric normalisation to 10.4 kg; prior evaluated at 3.19 kg)",
    "Prior only (clearance-maturation function; prior evaluated at 39 1/7 weeks)",
    "Screened and rejected (higher ELPD but diagnostic misspecification)",
    "Exposure-response figure stratifier only",
    "Exposure-response figure stratifier only"
  ),
  row.names = NULL
) |>
  knitr::kable(caption = "Demographics screened by Lam 2025 but not retained in the final model.")
```

# Population

The analysis cohort is 36 neonates with in-utero opioid exposure enrolled in
a substudy of NCT01965704, a double-blind, placebo-controlled, multicenter
trial in 98 mother/neonate dyads (Table 1). Mean (SD) gestational age at
birth was 38.6 (1.1) weeks, birth weight 3.1 (0.4) kg, and length at birth
49.7 (2.7) cm; the cohort was 18 male / 18 female. The median (min, max)
number of neonatal ondansetron doses received was 2 (1, 5) and the median
number of plasma samples per neonate was 3 (1, 5). The cohort median birth
weight and postmenstrual age used to build the informative priors were
3.19 kg and 39 1/7 weeks (Table 2 footnote).

Mothers randomised to ondansetron received a single 8 mg intravenous dose
within 4 h of delivery, repeated once if delivery had not occurred within
4 h (median 1, max 2 doses). Neonates received ondansetron 0.07 mg/kg orally
once every 24 h starting on the day of birth, up to five doses, with the
first dose given 4-8 h after delivery; the mean (SD) neonatal dose was
0.21 (0.03) mg. Maternal plasma ondansetron concentration at delivery was
28.2 (23.5) ng/mL across 29 maternal samples.

More than half of the neonates had measurable pre-dose ondansetron from
transplacental transfer. Lam 2025 handled this by initialising the neonatal
central compartment to the observed maternal concentration
(Appendix S1 `$PK`: `A_0(2) = C0 * S2`, with `S2 = V/1000`); the umbilical
cord was not carried as an intermediate compartment.

The same information is available programmatically via
`readModelDb("Lam_2025_ondansetron")()$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Lam_2025_ondansetron.R`. The
table below collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` (KA) | 0.19 /h | Table 3, KA row, posterior median (95% CDI 0.15, 0.23) |
| `lcl` (CL) | 0.58 L/h | Table 3, CL row, posterior median (95% CDI 0.51, 0.67) |
| `lvc` (V) | 0.29 L | Table 3, V row, posterior median (95% CDI 0.26, 0.32) |
| `lvp` (V2) | 0.91 L | Table 3, V2 row, posterior median (95% CDI 0.75, 1.11) |
| `lq` (Q) | 6.15 L/h | Table 3, Q row, posterior median (95% CDI 5.63, 6.74) |
| `lfdepot` (F) | 0.62, fixed | Table 3, F row `0.62 (FIXED)`; Methods 2.2 "Bioavailability (F) was fixed to 0.62 for all model runs" |
| `etalcl` | 0.54 (log-scale variance) | Table 3, "Interindividual Variability (% CV): CL" row, posterior median 0.54 (84.2% CV; 95% CDI 0.32, 0.92) |
| `addSd` | 8.99 ng/mL | Table 3, "Residual error: Additive error (SD)" row, posterior median 80.9 (95% CDI 59.3, 114.6); table note "SD of sigma = sqrt(posterior median)" |
| `d/dt(depot) = -ka * depot` | n/a | Appendix S1 `$DES`: `K12=KA`, `DADT(1)=-K12*A(1)` |
| `d/dt(central) = ka*depot - kel*central - k12*central + k21*peripheral1` | n/a | Appendix S1 `$DES`: `K20=CL/V`, `K23=CLD/V`, `K32=CLD/V2`, `DADT(2)=K12*A(1)-K20*A(2)-K23*A(2)+K32*A(3)` |
| `d/dt(peripheral1) = k12*central - k21*peripheral1` | n/a | Appendix S1 `$DES`: `DADT(3)=K23*A(2)-K32*A(3)` |
| `Cc = 1000 * central / vc` | n/a | Appendix S1 `$PK`: `S2=V/1000` (dose in mg, V in L, concentration in ng/mL) |
| `Cc ~ add(addSd)` | n/a | Methods 2.2, `Y_ij = Yhat_ij + eps_ij`; Appendix S1 `$ERROR`: `Y = F + ERR(1)` |
| `central(0) = CP0_MAT_NGML * vc / 1000` | n/a | Appendix S1 `$PK`: `A_0(2) = C0 * S2` with `S2 = V/1000`; Results 3.1 ("initializing the neonatal central compartment to the observed maternal concentrations") |
| `CP0_MAT_NGML` cohort distribution | 28.2 (23.5) ng/mL, 29 samples | Table 1, "Maternal plasma ondansetron concentration (ng/mL)" |
| beta_CL = 0.76, T_CL = 3.82 months (prior only) | fixed | Table 3 rows "B CL" and "T CL"; Methods 2.2.1 maturation function |
| Allometric exponents 0.75 (CL, Q) / 1 (V, V2), reference weight 10.4 kg (prior only) | fixed | Methods 2.2.1, `P = theta_P * (WT_i / 10.4)^theta_allo` |

# Virtual cohort

Original observed data are not publicly available. The cohorts below are
virtual populations whose demographics and dosing follow the published trial
description (Table 1, Methods 2.1). Two cohort families are simulated:

1. A **characterisation arm** - a single oral 0.07 mg/kg dose with no
   maternal transfer and dense sampling, used for a clean non-compartmental
   characterisation of the model (Cmax, Tmax, AUC0-inf, terminal half-life).
2. **Trial-replication arms** - neonates receiving 1, 2, 3 or 5 oral doses
   q24h with transplacental maternal transfer, integrated from birth to
   75 h after the last dose (the paper's `tl`), used to reproduce the
   cumulative AUC0-tl and Cmax distributions underlying Figures 3 and 4.

```{r cohort}
set.seed(20250211)

n_per_arm <- 120L
first_dose_time <- 6      # h; the trial gave the first neonatal dose 4-8 h after delivery
tau <- 24                 # h; q24h dosing
tl_lag <- 75              # h; the paper integrates to 75 h after the last dose
dose_per_kg <- 0.07       # mg/kg

# Pull the fixed effects straight out of the model so the vignette can never
# drift from the packaged ini() values.
theta_val <- function(nm) {
  d <- ui$iniDf
  v <- d$est[!is.na(d$ntheta) & d$name == nm]
  stopifnot(length(v) == 1L)
  unname(v)
}

# Typical central volume, used to convert the maternal plasma concentration
# into the amount that initialises the neonatal central compartment
# (Appendix S1: A_0(2) = C0 * S2 with S2 = V/1000). The model has no IIV on V,
# so this is exact for every subject.
vc_typ <- exp(theta_val("lvc"))
f_bioav <- exp(theta_val("lfdepot"))
c(vc_L = vc_typ, F = f_bioav)

rtnorm <- function(n, mean, sd, lower = -Inf, upper = Inf) {
  out <- stats::rnorm(n, mean, sd)
  bad <- out < lower | out > upper
  while (any(bad)) {
    out[bad] <- stats::rnorm(sum(bad), mean, sd)
    bad <- out < lower | out > upper
  }
  out
}

# Observation grid: fine near time 0 (the maternal-transfer bolus redistributes
# with a distribution half-life of a couple of minutes), fine after each oral
# dose (to resolve Cmax), and hourly out to tl.
make_grid <- function(dose_times, tl) {
  sort(unique(c(
    seq(0, 0.6, by = 0.01),
    seq(0.6, 4, by = 0.05),
    unlist(lapply(dose_times, function(d) d + seq(0, 12, by = 0.25))),
    seq(0, tl, by = 1),
    tl
  ))) |>
    (\(x) x[x <= tl])()
}

make_trial_arm <- function(n_doses, n, id_offset) {
  dose_times <- first_dose_time + tau * seq_len(n_doses) - tau
  tl <- max(dose_times) + tl_lag
  grid <- make_grid(dose_times, tl)

  subj <- tibble(
    id = id_offset + seq_len(n),
    WT_BIRTH = rtnorm(n, 3.1, 0.4, lower = 2.0, upper = 4.6),
    CP0_MAT_NGML = rtnorm(n, 28.2, 23.5, lower = 0, upper = 120),
    arm = paste0(n_doses, if (n_doses == 1) " dose" else " doses"),
    n_doses = n_doses
  )

  oral <- subj |>
    tidyr::crossing(time = dose_times) |>
    mutate(evid = 1L, cmt = "depot", amt = dose_per_kg * WT_BIRTH)

  obs <- subj |>
    tidyr::crossing(time = grid) |>
    mutate(evid = 0L, cmt = "central", amt = NA_real_)

  bind_rows(oral, obs) |>
    arrange(id, time, desc(evid))
}

trial_events <- bind_rows(
  make_trial_arm(1L, n_per_arm, id_offset =   0L),
  make_trial_arm(2L, n_per_arm, id_offset = 200L),
  make_trial_arm(3L, n_per_arm, id_offset = 400L),
  make_trial_arm(5L, n_per_arm, id_offset = 600L)
)
stopifnot(!anyDuplicated(unique(trial_events[, c("id", "time", "evid")])))

# Characterisation arm: single oral dose, no maternal transfer, dense grid.
# CP0_MAT_NGML = 0 is the "born with no drug on board" case, so the profile
# below is the contribution of the oral dose alone.
char_subj <- tibble(
  id = 1000L + seq_len(n_per_arm),
  WT_BIRTH = rtnorm(n_per_arm, 3.1, 0.4, lower = 2.0, upper = 4.6),
  CP0_MAT_NGML = 0,
  arm = "single dose (characterisation)"
)
char_events <- bind_rows(
  char_subj |> mutate(time = 0, evid = 1L, cmt = "depot",
                      amt = dose_per_kg * WT_BIRTH),
  char_subj |>
    tidyr::crossing(time = sort(unique(c(seq(0, 12, by = 0.1),
                                         seq(12, 48, by = 0.5))))) |>
    mutate(evid = 0L, cmt = "central", amt = NA_real_)
) |>
  arrange(id, time, desc(evid))

c(trial_rows = nrow(trial_events), char_rows = nrow(char_events))
```

# Simulation

`useLinCmt = FALSE` keeps rxode2 from auto-converting the explicit
three-state ODE system into an analytic `linCmt()` solution.

The tightened solver tolerances are load-bearing rather than cosmetic. This
model is stiff (`k12 = q/vc` is about 21 /h against `ka = 0.19 /h`) and the
84% CV on CL puts a few subjects at CL near 5 L/h, i.e. `kel` near 19 /h. The
paper's integration window runs to 75 h past the last dose, by which point
those subjects' concentrations are about 1e-9 of their Cmax. At rxode2's
default `atol = 1e-8` that tail is below the absolute tolerance and a handful
of records come back very slightly negative (about -5e-5 ng/mL); PKNCA then
returns `NA` for those subjects, and a single `NA` silently propagates into
the arm-level `median()`. Tightening to `atol = 1e-12` removes every negative
record across the whole cohort, and the assertions below fail loudly rather
than let a future solver change reintroduce the problem quietly.

```{r simulate}
sim_trial <- rxode2::rxSolve(
  mod, events = trial_events,
  keep = c("arm", "n_doses", "WT_BIRTH", "CP0_MAT_NGML"),
  useLinCmt = FALSE, atol = 1e-12, rtol = 1e-10
) |>
  as.data.frame()

sim_char <- rxode2::rxSolve(
  mod, events = char_events,
  keep = c("arm", "WT_BIRTH", "CP0_MAT_NGML"),
  useLinCmt = FALSE, atol = 1e-12, rtol = 1e-10
) |>
  as.data.frame()

# rxSolve silently drops subjects when an event table is malformed; assert the
# subject count survived the solve.
stopifnot(
  dplyr::n_distinct(sim_trial$id) == 4L * n_per_arm,
  dplyr::n_distinct(sim_char$id) == n_per_arm
)

# No negative or missing concentrations anywhere: a negative deep-tail record
# is the signature of the tolerance problem described above, and PKNCA turns
# it into a silent NA.
stopifnot(
  !anyNA(sim_trial$Cc), !anyNA(sim_char$Cc),
  min(sim_trial$Cc) >= 0, min(sim_char$Cc) >= 0
)
c(min_Cc_trial = min(sim_trial$Cc), min_Cc_char = min(sim_char$Cc))
```

The maternal-transfer initial condition must be visible in the very first
observation: because the model sets `central(0) <- CP0_MAT_NGML * vc / 1000`
and reports `Cc <- 1000 * central / vc`, the two conversions cancel and `Cc`
at time 0 must equal the subject's maternal concentration exactly. This is a
direct per-subject test of the `A_0(2) = C0 * S2` line in the paper's `$PK`
block, and it fails loudly if the covariate is dropped or the unit conversion
is wrong in either direction.

```{r check-maternal-transfer}
t0 <- sim_trial |> filter(time == 0)
max_abs_err <- max(abs(t0$Cc - t0$CP0_MAT_NGML))
c(subjects_checked = nrow(t0), max_abs_err = max_abs_err)
stopifnot(nrow(t0) == 4L * n_per_arm, max_abs_err < 1e-8)

# The characterisation arm carries no maternal transfer, so its time-zero
# concentration must be exactly zero.
stopifnot(all(sim_char$Cc[sim_char$time == 0] == 0))
```

# Model characterisation (single 0.07 mg/kg oral dose)

Not a published figure; the paper's main-text figures are exposure-response
plots and its concentration-time diagnostics are in Figures S1-S15. This
panel shows the shape the model implies for a single neonatal dose.

```{r figure-profile}
sim_char |>
  group_by(time) |>
  summarise(
    Q05 = quantile(Cc, 0.05), Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  labs(x = "Time after dose (h)", y = "Ondansetron (ng/mL)",
       title = "Single oral 0.07 mg/kg dose in a neonate",
       caption = "Median with 5th-95th percentile band, 120 virtual neonates.")
```

Absorption (KA = 0.19 /h) is far slower than disposition, so the profile is
flip-flop: the terminal slope is set by KA rather than by elimination.

```{r pknca-char}
char_nca <- sim_char |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, arm)

char_nca <- bind_rows(
  char_nca,
  char_nca |> distinct(id, arm) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, arm, time, .keep_all = TRUE) |>
  arrange(id, arm, time)

char_dose <- char_events |>
  filter(evid == 1) |>
  select(id, time, amt, arm)

char_conc_obj <- PKNCA::PKNCAconc(char_nca, Cc ~ time | arm + id,
                                  concu = "ng/mL", timeu = "h")
char_dose_obj <- PKNCA::PKNCAdose(char_dose, amt ~ time | arm + id,
                                  doseu = "mg")

char_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  char_conc_obj, char_dose_obj,
  intervals = data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                         aucinf.obs = TRUE, half.life = TRUE)
))

char_tbl <- as.data.frame(char_res$result) |>
  # pk.nca() also returns the lambda-z regression diagnostics that back the
  # half-life; keep the four requested parameters.
  filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  group_by(PPTESTCD) |>
  summarise(Median = median(PPORRES, na.rm = TRUE),
            P05 = quantile(PPORRES, 0.05, na.rm = TRUE),
            P95 = quantile(PPORRES, 0.95, na.rm = TRUE), .groups = "drop") |>
  mutate(Parameter = nlmixr2lib::ncaParamLabel(PPTESTCD)) |>
  select(Parameter, Median, P05, P95)

knitr::kable(
  char_tbl, digits = 2,
  caption = "Non-compartmental characterisation of a single oral 0.07 mg/kg neonatal dose (120 virtual neonates)."
)
```

## Structural identity: AUC0-inf equals F x Dose / CL

For a linear model with first-order input, the exact per-subject identity
`AUC0-inf = F * Dose / CL` holds regardless of the absorption rate. Comparing
the PKNCA result to the closed form is a per-subject check on the whole
pipeline (dosing units, the `f(depot)` assignment, and the `1000 *` unit
conversion in `Cc`), not a median-versus-median comparison.

```{r auc-identity}
cl_i <- sim_char |> distinct(id, cl) |> arrange(id)

analytic_char <- char_subj |>
  left_join(cl_i, by = "id") |>
  mutate(auc_analytic = 1000 * f_bioav * dose_per_kg * WT_BIRTH / cl)

auc_check <- as.data.frame(char_res$result) |>
  filter(PPTESTCD == "aucinf.obs") |>
  select(id, auc_pknca = PPORRES) |>
  left_join(analytic_char |> select(id, auc_analytic), by = "id") |>
  mutate(rel_err = abs(auc_pknca - auc_analytic) / auc_analytic)

c(f = f_bioav,
  max_rel_err = max(auc_check$rel_err),
  median_rel_err = median(auc_check$rel_err))
stopifnot(max(auc_check$rel_err) < 0.02)
```

# Trial replication: cumulative exposure by number of doses

Lam 2025 simulated, for each neonate, the cumulative AUC from birth to `tl`
(75 h after the last study dose) and the maximum concentration over the same
window, then plotted the change in mean Finnegan scores against exposure
quantiles stratified by the total number of doses received (Figures 3 and 4).
The PD layer is not a model in the paper - the exposure-response analysis is
graphical - so what is reproduced here is the exposure (boxplot) layer of
those figures.

```{r pknca-trial}
trial_nca <- sim_trial |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, arm)

trial_nca <- bind_rows(
  trial_nca,
  trial_nca |> distinct(id, arm) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, arm, time, .keep_all = TRUE) |>
  arrange(id, arm, time)

trial_dose <- trial_events |>
  filter(evid == 1) |>
  select(id, time, amt, arm)

trial_conc_obj <- PKNCA::PKNCAconc(trial_nca, Cc ~ time | arm + id,
                                   concu = "ng/mL", timeu = "h")
trial_dose_obj <- PKNCA::PKNCAdose(trial_dose, amt ~ time | arm + id,
                                   doseu = "mg")

trial_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  trial_conc_obj, trial_dose_obj,
  intervals = data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                         auclast = TRUE)
))

exposure <- as.data.frame(trial_res$result) |>
  filter(PPTESTCD %in% c("cmax", "auclast")) |>
  select(arm, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  left_join(trial_events |> distinct(id, n_doses), by = "id")

# The paper's Cmax is the maximum from BIRTH to tl, so for a neonate whose
# maternal-transfer concentration exceeds every post-dose peak it is simply
# C0. Carry a post-first-dose maximum alongside it so the two contributions
# can be told apart in the comparison below.
cmax_postdose <- sim_trial |>
  filter(time >= first_dose_time) |>
  group_by(id) |>
  summarise(cmax_postdose = max(Cc), .groups = "drop")

exposure <- exposure |> left_join(cmax_postdose, by = "id")

# Every subject must have both metrics; an NA here would silently turn the
# arm-level medians and the attainment percentages below into NA.
stopifnot(
  nrow(exposure) == 4L * n_per_arm,
  !anyNA(exposure$auclast), !anyNA(exposure$cmax)
)
```

```{r figure-3-4-exposure, fig.height = 3.6}
exposure |>
  tidyr::pivot_longer(c(auclast, cmax), names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric, levels = c("auclast", "cmax"),
                         labels = c("Cumulative AUC0-tl (ng*h/mL)",
                                    "Cmax (ng/mL)"))) |>
  ggplot(aes(x = factor(n_doses), y = value)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~metric, scales = "free_y") +
  scale_y_log10() +
  labs(x = "Total number of ondansetron doses received",
       y = NULL,
       title = "Exposure by number of doses",
       caption = paste("Replicates the exposure (boxplot) layer of Figures 3 and 4",
                       "of Lam 2025; the Finnegan-score layer is not a model."))
```

## Comparison against published exposure landmarks

Lam 2025 publishes no NCA table. The quantitative exposure values it does
report are the medians of the exposure quantiles shown in Figures 1-4, quoted
in the Results text. The most completely specified stratum is the five-dose
group, for which both metrics are given: a change in mean Finnegan scores of
-3.59 units "between a median of 700 and 1650 ng*h/mL" (Figure 3) and
"between a median of 19.6 and 37.5 ng/mL" for Cmax (Figure 4). The reference
values below are the **geometric midpoints of those reported quantile-median
intervals**, which is the closest single-number summary the paper supports;
the interval itself is quoted alongside so a reader can see the width of the
target.

```{r nca_compare}
published <- tibble::tribble(
  ~arm,      ~auclast, ~cmax,
  "5 doses", sqrt(700 * 1650), sqrt(19.6 * 37.5)
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = exposure |> filter(arm == "5 doses") |>
    tidyr::pivot_longer(c(auclast, cmax), names_to = "PPTESTCD",
                        values_to = "PPORRES") |>
    as.data.frame(),
  reference = published,
  by = "arm",
  units = c(auclast = "ng*h/mL", cmax = "ng/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp, digits = 1,
  caption = paste(
    "Simulated (median of 120 virtual neonates) vs. the geometric midpoint of",
    "the quantile-median intervals reported by Lam 2025 for the five-dose",
    "stratum: AUC0-tl 700-1650 ng*h/mL (Figure 3), Cmax 19.6-37.5 ng/mL",
    "(Figure 4). * differs from reference by >20%."
  )
)
```

The AUC row is the informative one. Cmax carries a known upward bias in this
reproduction: the paper defines Cmax as the maximum from **birth**, so for a
neonate whose maternal-transfer concentration exceeds every post-dose peak it
is simply `C0`, and the virtual cohort assigns a maternal concentration drawn
from N(28.2, 23.5) ng/mL to **every** subject, whereas the trial reports that
only "more than half" of neonates had a measurable pre-dose concentration.
Separating the two contributions shows the size of that effect:

```{r cmax-decomposition}
exposure |>
  left_join(trial_events |> distinct(id, CP0_MAT_NGML), by = "id") |>
  group_by(arm) |>
  summarise(
    `Median maternal CP0_MAT_NGML (ng/mL)` = median(CP0_MAT_NGML),
    `Median Cmax from birth (ng/mL)` = median(cmax),
    `Median Cmax after the first dose (ng/mL)` = median(cmax_postdose),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 1,
               caption = "Cmax from birth (the paper's definition, which includes the maternal-transfer spike) vs. the post-dose maximum.")
```

```{r landmark-brackets}
exposure |>
  group_by(arm) |>
  summarise(
    `Median AUC0-tl (ng*h/mL)` = median(auclast),
    `Median Cmax (ng/mL)` = median(cmax),
    .groups = "drop"
  ) |>
  mutate(
    `Lam 2025 reported quantile-median interval` = c(
      "not reported", "not reported", "not reported",
      "AUC 700-1650 ng*h/mL; Cmax 19.6-37.5 ng/mL (Figures 3, 4)"
    )
  ) |>
  knitr::kable(
    digits = 1,
    caption = "Simulated median exposure by dose-count arm, with the published landmark where one exists."
  )
```

The whole-cohort landmarks reported elsewhere in the paper bracket the
simulated values in the same way: Figure 2 quotes Cmax quantile medians of
19.6-37 ng/mL for the ondansetron-plus-morphine stratum, Figure S17 quotes
53.2-73.1 ng/mL for term neonates, and Figure 1 quotes cumulative AUC
quantile medians of about 645 ng*h/mL (ondansetron plus morphine) and about
1420 ng*h/mL (ondansetron only).

## Reproducing the "at least three doses" conclusion

The paper's headline dosing conclusion is that "an AUC0-tl of at least
1000 ng*h/mL appears to be the threshold for reducing mean Finnegan scores"
and that "at least three doses of the current oral ondansetron regimen are
required to reduce symptom severity". Those two statements are jointly
testable against the packaged model: with a typical CL the median neonate
does not reach 1000 ng*h/mL until four or five doses, but the large
inter-individual variability in CL (84% CV) means a meaningful share of
three-dose neonates does.

```{r threshold-attainment}
exposure |>
  group_by(arm) |>
  summarise(
    `Median AUC0-tl (ng*h/mL)` = median(auclast),
    `% reaching AUC0-tl >= 1000 ng*h/mL` = 100 * mean(auclast >= 1000),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 1,
               caption = "Attainment of the 1000 ng*h/mL cumulative-AUC landmark by number of doses.")
```

```{r threshold-monotone}
attain <- exposure |>
  group_by(n_doses) |>
  summarise(pct = 100 * mean(auclast >= 1000), .groups = "drop") |>
  arrange(n_doses)
attain
# Attainment must increase monotonically with the number of doses, and the
# one-dose arm must be far below the three-dose arm.
stopifnot(
  all(diff(attain$pct) > 0),
  attain$pct[attain$n_doses == 1] < 10,
  attain$pct[attain$n_doses == 3] > 20
)
```

## Structural identity for the cumulative AUC

The paper's AUC0-tl is the integral of `A_central / V` from birth to `tl`.
Because `tl` sits many terminal half-lives after the last dose, that integral
is numerically the total AUC, which for this linear model has the exact
closed form `(A_maternal + n_doses * F * Dose) / CL`. Checking PKNCA's
trapezoidal AUC against the closed form per subject validates the maternal
initial condition and the multi-dose bookkeeping together.

```{r auc-identity-trial}
cl_trial <- sim_trial |> distinct(id, cl)

auc_check_trial <- exposure |>
  left_join(trial_events |> distinct(id, WT_BIRTH, CP0_MAT_NGML), by = "id") |>
  left_join(cl_trial, by = "id") |>
  mutate(
    amt_total = CP0_MAT_NGML * vc_typ / 1000 +
      n_doses * f_bioav * dose_per_kg * WT_BIRTH,
    auc_analytic = 1000 * amt_total / cl,
    rel_err = abs(auclast - auc_analytic) / auc_analytic
  )

auc_check_trial |>
  group_by(arm) |>
  summarise(`Median relative error` = median(rel_err),
            `Max relative error` = max(rel_err), .groups = "drop") |>
  knitr::kable(digits = 4,
               caption = "PKNCA trapezoidal AUC0-tl vs. the closed-form (A_maternal + n * F * Dose) / CL.")

stopifnot(max(auc_check_trial$rel_err) < 0.03)
```

# Assumptions and deviations

* **Bioavailability is applied to the depot.** Methods 2.2 states that
  "Bioavailability (F) was fixed to 0.62 for all model runs" and Table 3
  reports `F 0.62 (FIXED)`, so the packaged model applies
  `f(depot) <- 0.62`. The Appendix S1 control stream agrees on the *value*
  but not on its use: it declares `$THETA ... 0.49 FIX ;F`, and 0.49 is
  `logit(0.62) = log(0.62 / 0.38) = 0.4895` (the five structural thetas in
  that block are log-transformed, as its `; log transformed thetas` header
  says, but `exp(0.49) = 1.63` is not a possible bioavailability, so the F
  theta must be logit-scaled). What the printed stream lacks is any
  `F1 = ...` assignment in `$PK` referencing that theta, so as listed it
  declares a fixed F and never applies it - most likely an omission from the
  appendix listing. The paper's prose and parameter table are taken as
  authoritative, and the published cumulative AUC independently
  discriminates the two readings. For a typical neonate (median CL 0.58 L/h, birth weight
  3.19 kg) the closed-form cumulative AUC0-tl after five doses is
  `1000 * (A_maternal + 5 * F * 0.07 * 3.19) / 0.58`: with F = 0.62 that
  is about 1210 ng*h/mL, which falls inside the 700-1650 ng*h/mL
  quantile-median interval Lam 2025 reports for the five-dose stratum
  (Figure 3); with F = 1 it is about 1940 ng*h/mL, roughly 18% above the
  top of that interval and 80% above its geometric midpoint. The simulated
  cohort above bears this out: the five-dose median cumulative AUC0-tl comes
  out within 0.2% of the 1070 ng*h/mL geometric midpoint of the published
  interval, whereas F = 1 would scale it by
  `(A_maternal + 5 * 0.07 * WT) / (A_maternal + 5 * 0.62 * 0.07 * WT)`, about
  1.61x, to roughly 1720 ng*h/mL - above the top of the published interval and
  about 61% above its midpoint. Cmax does not discriminate (typical post-dose
  peak 24 ng/mL at F = 0.62 vs 39 ng/mL at F = 1, both inside the
  19.6-73.1 ng/mL span of the published Cmax quantile medians), so the AUC
  comparison is what settles it.
* **Additive residual SD taken as sqrt(80.9) = 8.99 ng/mL.** Table 3
  reports the sigma posterior median as 80.9 with a parenthetical SD of
  9.2, and the table note defines "SD of sigma = sqrt(posterior median)".
  Applying the paper's own relationship to the printed median gives 8.99,
  a 2% disagreement with the printed 9.2. The value derived from the
  estimated quantity (the posterior median of the variance) is used. The
  analogous IIV row is self-consistent to rounding: the note's formula
  `%CV = sqrt(exp(median) - 1) * 100` applied to the printed 0.54 gives
  84.6% against the printed 84.2%.
* **No individual-level covariates.** Allometric scaling to 10.4 kg with
  fixed exponents (0.75 for CL and Q, 1 for V and V2) and the
  clearance-maturation term
  `1 - beta_CL * exp(-(AGE - 1) * ln(2) / T_CL)` with beta_CL = 0.76 and
  T_CL = 3.82 months are properties of the infant reference model. Lam 2025
  used them once, at the cohort median birth weight and postmenstrual age,
  to compute the priors; the selected final model applies no covariate
  effect at the individual level (Results 3.1; Appendix S1 `$PK`). The
  packaged model therefore has no covariates, and beta_CL / T_CL are
  documented rather than encoded as `ini()` parameters, because encoding a
  parameter that the model body never uses would misrepresent the fit.
* **The infant reference model itself is not packaged.** Lam 2025 cites it
  (reference 17) and reproduces its equations, but not its parameter values
  at the 10.4 kg reference weight, so it cannot be reconstructed from this
  paper alone.
* **Maternal transfer is encoded as a state initial condition, driven by a
  covariate.** The paper's `A_0(2) = C0 * S2` sets the neonatal central
  compartment to `C0 * V / 1000` mg at birth, and it lives in the `$PK` block
  -- it is part of the model, not a feature of the dataset. The packaged model
  reproduces it literally as
  `central(0) <- CP0_MAT_NGML * vc / 1000`, with `CP0_MAT_NGML` (the mother's
  observed plasma concentration at delivery, ng/mL) declared in
  `covariateData` and newly registered in
  `inst/references/covariate-columns.md`. Two consequences a downstream user
  should know. First, `CP0_MAT_NGML` is a **required** covariate column: set
  it to 0 for a neonate with no transplacental transfer, as the
  characterisation arm above does. Second, because the initial condition is
  assigned inside `model()`, it takes precedence over any `inits =` passed to
  `rxSolve()`; that is the behaviour of the published model and matches how
  NONMEM's `A_0(2) =` works. Delivering the same amount as a time-zero
  `evid = 1` bolus into `central` would be numerically equivalent, but it
  would push a structural feature of the paper's model out of the model file
  and into every user's event table.
* **Distributional assumptions for the virtual cohort.** Birth weight is
  drawn as N(3.1, 0.4) kg truncated to 2.0-4.6 kg and the maternal plasma
  concentration as N(28.2, 23.5) ng/mL truncated to 0-120 ng/mL, matching
  the means and SDs in Table 1; the paper reports no distributional shape
  or per-subject values. In particular, every virtual neonate is given a
  maternal-transfer concentration, whereas the trial reports that only
  "more than half" of neonates had a measurable pre-dose concentration
  (29 maternal samples for 36 neonates); this biases the simulated Cmax
  from birth upward and is the reason that row is starred in the
  comparison table. The first neonatal dose is placed at 6 h, the midpoint
  of the protocol's 4-8 h window.
* **The exposure-response layer is not a model.** Lam 2025's Figures 1-4
  plot average change in mean Finnegan scores against exposure quantiles;
  no PD model, hazard model or covariate-effect equation is estimated for
  that relationship, so nothing is packaged for it. Only the exposure
  (boxplot) layer of Figures 3 and 4 is reproduced here.
* **Quantile-median references are not an NCA table.** The reference values
  in the comparison table are geometric midpoints of the quantile-median
  intervals quoted in the Results text for the five-dose stratum, because
  the paper publishes no NCA summary. A direct quantile-by-quantile
  comparison would be misleading: the paper's per-neonate exposures come
  from shrunk empirical-Bayes CL estimates on 1-5 samples per neonate, so
  their observed spread is much narrower than the estimated 84% CV
  population variability that the packaged model simulates.

# Errata

No erratum or correction notice for Lam 2025 was found on the journal
landing page, in PubMed, or in the PMC record (checked 2026-08-14). Two
internal inconsistencies in the article itself are recorded above and
handled explicitly: the Appendix S1 control stream never applies the
bioavailability theta it declares, and Table 3's parenthetical residual-error
SD (9.2) does not equal the square root of the posterior median it is
printed beside (sqrt(80.9) = 8.99).
