---
title: "Vildagliptin GLP-1 / glucose / insulin systems model (Landersdorfer 2012)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vildagliptin GLP-1 / glucose / insulin systems model (Landersdorfer 2012)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.4)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
mod <- readModelDb("Landersdorfer_2012_vildagliptin_qsp")
ui  <- rxode2::rxode(mod)
```

* Citation: `r ui$reference`
* Article: <https://doi.org/10.1111/j.1365-2125.2011.04109.x> (PMC3370342, open access)
* Companion PK / DPP-4 article: <https://doi.org/10.1111/j.1365-2125.2011.04108.x>
  (packaged separately as `modellib("Landersdorfer_2012_vildagliptin")`)

This is the pharmacodynamic half of a companion pair. The companion article
describes vildagliptin PK and DPP-4 activity as a target-mediated drug
disposition (TMDD) model; the present article adds three downstream endpoints
-- active GLP-1, plasma glucose and plasma insulin -- and states that "in the
full PK/PD model all PK (shown in the companion report) and PD parameters were
estimated at the same time" (Methods, p. 376). The whole system is therefore
one jointly-fitted model and is packaged as one file.

# Population

`r ui$population$n_subjects` adults with type 2 diabetes mellitus diagnosed at
least three months before screening, after a washout from hypoglycaemic drugs
of up to four weeks. Mean (range) weight 91 (65-116) kg, height 166 (148-183)
cm, age 53.5 (37-64) years; seven female and six male (Results, first
paragraph). The design was a randomized, double-blind, placebo-controlled
four-way crossover: oral vildagliptin 10, 25 or 100 mg twice daily for 28 days,
plus placebo. Twelve subjects completed all four periods and one completed only
the 10 and 25 mg treatments. Patients were confined from the evening of day 26
to the morning of day 29 of each period and received a standard diet with
identical meals across all four treatments (55% carbohydrate, 25% fat, 20%
protein). GLP-1, glucose and insulin were sampled on day 28.

The same information is available programmatically via
`readModelDb("Landersdorfer_2012_vildagliptin_qsp")()$population`.

# Errata and non-paper-derived values

Two quantities the model needs are not usable as printed. Both are resolved
below from the paper's own figures; **read this section before using the
model.**

## 1. `V_glc` is not reported anywhere

The glucose ODE divides the gut absorption rate by a volume of distribution of
glucose,
`Glc_GutAb = (k_aB*A_GB + k_aL*A_GL + k_aD*A_GD + k_aS*A_GS) / V_glc`.
The Methods define the symbol -- "where Vglc (dl) is the volume of distribution
of glucose" (p. 376) -- and then never give a value: it is absent from Table 1,
from the body text, from Figure 1, from the companion article, and there is no
supplement and no erratum.

It does not cancel out. The meal-amount scale is pinned independently by the
GLP-1 arm (`S1` is reported in g^-1 and multiplies `Glc_Gut`), so `V_glc` is a
free constant that sets the magnitude of every glucose excursion and, through
the glucose-insulin feedback, the insulin profile too.

Per operator ruling (2026-09-02) it was **back-solved from the Figure 3 placebo
panel**: the published median curve was digitised and `V_glc` fitted to it,
giving **250 dL (25.0 L)**. Ninety-six combinations of the unreported meal
times and food-intake durations bracket the estimate at 246-255 dL, so the
value is set by the glucose *level* rather than by the timing assumptions.

Two independent consequences, neither used in the fit, support it:

* endogenous glucose production `B_glc * k_out_glc * V_glc` = 11.1 g/h
  = 2.0 mg/kg/min at the cohort's mean 91 kg, the textbook human value;
* the resulting typical-value profile also reproduces Figure 7A-D (below),
  which the fit never saw.

The glucose validation against Figure 3 is therefore **partly circular** and is
labelled as such below; the Figure 2, 4 and 7 comparisons are not.

## 2. `S5` is printed with a lost unit prefix

Table 1 gives `S5 = 0.584` with units `l mIU-1`. Taken literally the model
cannot reproduce any of the paper's own figures -- glucose elimination at a
postprandial peak would be roughly 40 h^-1, a glucose half-life near one
minute -- and the best-fitting `V_glc` becomes 2 dL, a physiological
impossibility.

Figure 7B settles it directly. That panel plots
`ST_ins = S5 * [1 + S4 * (C_glp - B_glp)]` against time on a **0 to 0.009**
axis, and at baseline GLP-1 the bracket is exactly 1, so the panel reads `S5`
off its own y-axis. The published `0.584 l mIU-1` is **640-fold above the top
of that axis**; `5.84e-4 l mIU-1` = `0.584 mL/mIU` sits on the placebo curve's
fasting value of about 0.001. The paper drops a unit prefix on this pair of
quantities elsewhere too: the Results section (p. 377) prints `ST_glc` as
`l mg-1` where Table 1 and dimensional analysis both require `dl mg-1`.

The model file keeps the printed `0.584` in `ini()` and applies an explicit
`mL -> L` conversion in `model()`, so the published number stays visible and
the correction is auditable on one line.

No other parameter needed adjustment: Figure 7C confirms `S3 = 0.0185 dl mg-1`
on its own axis, Figure 7A confirms `R_maxC * cf2 = 3.2 h^-1`, and the
Discussion's "increase in insulin secretion of 35% for a 5 pmol l-1 increase in
GLP-1" reproduces exactly from `S2` (checked below).

# Source trace

Every `ini()` entry carries an in-file comment naming its source location; the
table collects them.

| Parameter / equation | Value | Source location |
|---|---|---|
| `lka1`, `lka2`, `lcl`, `lvc`, `lvp`, `lq` | 1.26, 1.05 /h; 36.4 L/h; 22.2, 97.3 L; 40.1 L/h | Companion (04108.x) Table 1 |
| `lfdepot` | 0.772 (fixed) | Companion Table 1, footnote * |
| `lkd`, `lk2`, `lkoff`, `lkdeg` | 71.9 nmol/L; 23.4, 0.612, 0.110 /h | Companion Table 1 |
| `lrmaxc`, `lrmaxp`, `lcf1` | 5.0, 13000 nmol; 2.80 | Companion Table 1 (`R_maxP` printed "13.0 mmol"; see below) |
| `lkout_glp_lin`, `ls1`, `lcf2`, `lbl_glp` | 2.07 /h; 0.049 /g; 0.641; 1.68 pmol/L | Table 1, p. 380 |
| `lfdepot_breakfast/lunch/dinner/snack` | 0.796, 0.865, 0.817, 0.342 | Table 1 (`F_B`, `F_L`, `F_D`, `F_S`) |
| `lka_breakfast/lunch/dinner/snack` | 0.732, 0.520, 0.252, 0.169 /h | Table 1 (`k_aB`, `k_aL`, `k_aD`, `k_aS`) |
| `lbl_glc`, `lkout_glc`, `ls5`, `ls4` | 133 mg/dL; 0.334 /h; 0.584 mL/mIU; 1.90 L/pmol | Table 1 (`S5` prefix per Figure 7B) |
| `lv_glc` | 250 dL | **Not published** -- back-solved from Figure 3 |
| `lbl_ins`, `lkout_ins`, `ls3`, `ls2` | 9.75 mIU/L; 14.0 /h; 0.0185 dL/mg; 0.0701 L/pmol | Table 1 |
| IIV (all etas) | BSV% squared | Table 1 "BSV (%)"; S-ADAPT reports sqrt(variance) (Methods p. 376) |
| Residual error (5 outputs) | `CV`/`SD` pairs | Table 1 (PD) and companion Table 1 (PK) |
| Meal gut ODEs, `Glc_Gut` | n/a | Methods "Structural PD model", p. 375 |
| GLP-1 ODE, `k_in_glp` steady state | n/a | Methods p. 376 |
| `Glc_GutAb`, glucose ODE, insulin ODE, `ST_ins`, `ST_glc` | n/a | Methods p. 376 |

Two printed-unit problems were resolved from internal evidence and are recorded
in the model file next to the affected line:

* `S1` -- Table 1 gives `g-1`, the body text (p. 376) gives `mg-1`. Table 1 is
  right: with `S1 = 0.049 mg-1` a ~50 g meal bolus would stimulate GLP-1
  secretion ~2450-fold. Gut amounts are therefore carried in **grams**.
* `R_maxP` -- the companion Table 1 prints "13.0 mmol"; 13 umol / 5 nmol
  = 2600-fold matches that paper's own ">2000-fold" tissue:plasma statement,
  whereas 13 mmol would be 2.6e6-fold and about 1.4 kg of a 110 kDa enzyme.

# Reconstructing the study day

The paper gives the meal structure but not the full timetable. What is stated:
meals were standardised and identical across periods; "breakfast and dinner
were consumed at approximately 30 min after the doses" (Methods, p. 374); the
evening dose was at 10.5 h (from the sampling schedule); and the input duration
`tk0` was "the actual recorded duration of food intake for each individual
patient and each meal", which is never reported. Lunch and snack times are not
given.

Lunch and snack are recovered from the published figures: Figures 2, 3, 4 and 7
all show four meal-driven excursions per day, spaced about 5.1, 5.4 and 3.5 h
apart. The schedule below combines the stated design with those spacings, and
`tk0` is assumed to be 45 min. `V_glc` moves by less than 4% across the whole
grid of plausible alternatives, so nothing downstream hinges on it.

```{r schedule}
MEAL_TIMES <- c(breakfast = 0.5, lunch = 5.6, dinner = 11.0, snack = 14.5)
TK0        <- 0.75   # h; duration of food intake (assumed -- not reported)
MEAL_G     <- 75     # g; the paper's arbitrary 75000 mg per-meal anchor
DOSE_TIMES <- c(0, 10.5)
N_DAY      <- 5L     # days of BID dosing simulated before the observed day
ARMS       <- c(Placebo = 0, `10 mg` = 10, `25 mg` = 25, `100 mg` = 100)
```

```{r events}
# One arm's event table. Observation records name an ENDPOINT (Cglc). This
# model declares five endpoints, so rxode2 requires observation rows to select
# among the endpoint slots; naming one does NOT renumber the ODE states (they
# keep slots 1-13) and every observable still comes back as its own column.
# The same idiom is used by the sibling Landersdorfer_2012_vildagliptin and by
# Bosch_2024_cotadutide_qsp vignettes.
# `obs` is a parameter rather than hard-coded because observation times MUST be
# added while the table is still an `rxEt`. Passing an already-materialized
# data.frame back into `rxode2::et()` and asking for extra sampling times
# silently DROPS them -- no error, no warning -- which makes any downstream
# `x[time %in% ...]` selection empty and any `stopifnot()` over it pass
# vacuously.
build_arm <- function(dose_mg, id,
                      obs = seq(24 * (N_DAY - 1L), 24 * N_DAY, by = 0.05)) {
  ev <- rxode2::et(id = id)
  for (d in seq_len(N_DAY) - 1L) {
    if (dose_mg > 0) {
      for (dt in DOSE_TIMES) {
        ev <- rxode2::et(ev, id = id, amt = dose_mg, cmt = "depot", time = 24 * d + dt)
      }
    }
    for (m in names(MEAL_TIMES)) {
      ev <- rxode2::et(ev, id = id, amt = MEAL_G, dur = TK0,
                       cmt = paste0("glucose_gut_", m), time = MEAL_TIMES[[m]] + 24 * d)
    }
  }
  rxode2::et(ev, id = id, obs, cmt = "Cglc")
}

events <- dplyr::bind_rows(lapply(seq_along(ARMS), function(i) {
  x <- as.data.frame(build_arm(ARMS[[i]], i))
  x$arm <- factor(names(ARMS)[i], levels = names(ARMS))
  x
}))
# No `unique()` here: de-duplicating first and then asking whether duplicates
# remain is always TRUE. Test the event table as actually built.
stopifnot(
  nrow(events) > 0,
  !anyDuplicated(events[, c("id", "time", "evid", "cmt")])
)
```

# Typical-value simulation

The published Figure 7 panels are typical-value simulations of model subparts,
so the primary validation is run with the random effects zeroed.

```{r simulate-typical}
sim <- rxode2::rxSolve(
  rxode2::zeroRe(mod), events = events, keep = "arm",
  returnType = "data.frame",
  # rxode2's ODE -> linCmt auto-conversion corrupts the endpoint mapping for
  # this multi-output model (known-vignette-failure-patterns.md pattern 5b).
  useLinCmt = FALSE
) |>
  dplyr::mutate(trel = time - 24 * (N_DAY - 1L))

stopifnot(nrow(sim) > 0, !anyNA(sim$Cglc), !anyNA(sim$Cglp), !anyNA(sim$Cins))

# Guard for the `cmt = "Cglc"` observation records above. Naming an ALGEBRAIC
# observable as a compartment is the classic slot-renumbering bug; naming a
# DECLARED ENDPOINT is not, because the endpoint slots already exist after the
# 13 ODE states. This model has five endpoints, so rxode2 rejects both
# `cmt = "central"` and an absent cmt on observation rows -- an endpoint must
# be selected. These assertions prove no renumbering occurred: all thirteen
# ODE states come back, all five observables come back, and the 100 mg dose
# lands in `depot` at exactly amt * F * 1e6 / MW nmol.
ode_states <- c("depot", "transit1", "central", "peripheral1", "complex",
                "complex_peripheral", "glucose_gut_breakfast", "glucose_gut_lunch",
                "glucose_gut_dinner", "glucose_gut_snack", "glp1", "glucose", "insulin")
stopifnot(
  all(ode_states %in% names(sim)),
  all(c("Cc", "DPP4", "Cglp", "Cglc", "Cins") %in% names(sim))
)
dose_check <- rxode2::rxSolve(
  rxode2::zeroRe(mod),
  rxode2::et(rxode2::et(amt = 100, cmt = "depot", time = 0), 1e-6, cmt = "Cglc"),
  returnType = "data.frame", useLinCmt = FALSE
)
stopifnot(abs(max(dose_check$depot) - 100 * 0.772 * 1e6 / 303.40) < 1)

# Day-to-day stability: the observed day must be at steady state. The
# observation grid is widened to cover days 4 and 5 (see the note in
# `build_arm` -- the landmark times have to be requested while the table is
# still an `rxEt`).
ss <- rxode2::rxSolve(
  rxode2::zeroRe(mod),
  build_arm(100, 1L, obs = seq(0, 24 * N_DAY, by = 0.05)),
  returnType = "data.frame", useLinCmt = FALSE
)
# 06:00 on each of days 1..5.
ss_t <- 24 * (seq_len(N_DAY) - 1L) + 6
ss_i <- vapply(ss_t, function(tt) which.min(abs(ss$time - tt)), integer(1))
# Guard the selection BEFORE using it: an empty or mismatched match would make
# every assertion below vacuously true rather than failing loudly.
stopifnot(
  length(ss_i) == N_DAY,
  all(abs(ss$time[ss_i] - ss_t) < 1e-6),
  all(is.finite(ss$Cglc[ss_i]))
)
ss_glc  <- ss$Cglc[ss_i]
ss_pct  <- 100 * abs(ss_glc[[N_DAY]] - ss_glc[[N_DAY - 1L]]) / ss_glc[[N_DAY - 1L]]
# Day-1 -> day-5 drift, so the reader can see the metric is capable of moving
# and that its smallness above is convergence rather than a degenerate compare.
ss_drift <- 100 * abs(ss_glc[[N_DAY]] - ss_glc[[1L]]) / ss_glc[[1L]]
stopifnot(
  length(ss_pct) == 1L, is.finite(ss_pct), ss_pct < 1,
  # the approach to steady state must be monotone and must actually happen
  ss_drift > 100 * ss_pct, ss_drift > 0.5
)
```

Day 4 and day 5 glucose at 06:00 differ by `r sprintf("%.1e%%", ss_pct)`, so the
profiles below are at steady state. The metric is not degenerate: the same
quantity moves `r sprintf("%.2f%%", ss_drift)` (from
`r sprintf("%.1f", ss_glc[[1]])` to `r sprintf("%.1f", ss_glc[[N_DAY]])` mg/dL)
between day 1 and day 5, so it is convergence that makes the day-4-to-day-5
figure small, not a comparison of a value with itself.

# Replicating Figure 3 (glucose)

The placebo panel of Figure 3 was digitised (median curve, 1208 points) and is
shown here at eight landmark times. This is the curve `V_glc` was fitted to, so
agreement is expected -- it is reported as a goodness-of-back-solve, not as
independent validation.

```{r figure-3}
fig3_placebo <- tibble::tribble(
  ~trel,  ~Cglc_pub, ~landmark,
  -3.73,  189.8,     "start of window",
  -0.51,  170.0,     "pre-breakfast",
   2.12,  265.7,     "breakfast peak",
   5.26,  219.5,     "pre-lunch trough",
   7.40,  278.9,     "lunch peak",
  10.37,  227.2,     "pre-dinner trough",
  12.14,  245.9,     "dinner peak",
  16.41,  225.0,     "end of window"
)

pb <- dplyr::filter(sim, arm == "Placebo")
fig3 <- fig3_placebo |>
  dplyr::mutate(
    Cglc_sim = approx(pb$trel, pb$Cglc, xout = trel, rule = 2)$y,
    pct_diff = 100 * (Cglc_sim - Cglc_pub) / Cglc_pub
  )

fig3 |>
  dplyr::mutate(dplyr::across(c(Cglc_pub, Cglc_sim, pct_diff), \(x) round(x, 1))) |>
  dplyr::rename(
    "Time after morning dose (h)" = trel,
    "Landmark"                    = landmark,
    "Figure 3 median (mg/dL)"     = Cglc_pub,
    "Simulated (mg/dL)"           = Cglc_sim,
    "Difference (%)"              = pct_diff
  ) |>
  knitr::kable(caption = "Placebo glucose: digitised Figure 3 median vs typical-value simulation.")

ggplot(pb, aes(trel, Cglc)) +
  geom_line(linewidth = 0.8) +
  geom_point(data = fig3_placebo, aes(trel, Cglc_pub), size = 2.2, shape = 21, fill = "white") +
  labs(x = "Time after morning dose (h)", y = "Glucose (mg/dL)",
       title = "Figure 3, placebo panel",
       caption = "Line: packaged model, typical values. Points: digitised published median.")
```

```{r figure-3-gate}
# Deterministic comparison (zeroRe), so a tight bound is appropriate; the
# headroom covers the +/- 2 mg/dL digitisation error of the published curve.
stopifnot(
  stats::median(abs(fig3$pct_diff)) < 6,
  max(abs(fig3$pct_diff)) < 15
)
```

# Replicating Figures 2 and 4 (GLP-1 and insulin)

Neither of these was used to derive `V_glc` or the `S5` prefix, so they are
independent checks. Landmark values were read off the published placebo panels
by eye and carry roughly +/- 10% reading error, so the comparison is reported
as an order-of-magnitude and shape check.

```{r figure-2-4}
fig24 <- tibble::tribble(
  ~endpoint,          ~trel, ~published,
  "GLP-1 (pmol/L)",   -0.51,  2.7,
  "GLP-1 (pmol/L)",    2.12,  5.8,
  "GLP-1 (pmol/L)",    7.40,  5.6,
  "GLP-1 (pmol/L)",   12.14,  5.5,
  "Insulin (mIU/L)",  -0.51, 22.0,
  "Insulin (mIU/L)",   2.12, 50.0,
  "Insulin (mIU/L)",   7.40, 57.0,
  "Insulin (mIU/L)",  12.14, 45.0
) |>
  dplyr::mutate(
    simulated = ifelse(
      endpoint == "GLP-1 (pmol/L)",
      approx(pb$trel, pb$Cglp, xout = trel, rule = 2)$y,
      approx(pb$trel, pb$Cins, xout = trel, rule = 2)$y
    ),
    ratio = simulated / published
  )

fig24 |>
  dplyr::mutate(dplyr::across(c(published, simulated, ratio), \(x) round(x, 2))) |>
  dplyr::rename(
    "Endpoint"                    = endpoint,
    "Time after morning dose (h)" = trel,
    "Published median (read)"     = published,
    "Simulated"                   = simulated,
    "Simulated / published"       = ratio
  ) |>
  knitr::kable(caption = "Placebo GLP-1 (Figure 2) and insulin (Figure 4) landmarks.")
```

```{r figure-2-4-gate}
# Deterministic, but compared against eye-read landmarks: assert the right
# order of magnitude and no systematic inversion, not agreement to a few
# percent. A mis-transcribed baseline, rate constant or unit moves these by
# factors of 3 or more and still breaks the bound.
base_glp <- fig24$simulated[fig24$endpoint == "GLP-1 (pmol/L)"  & fig24$trel == -0.51]
base_ins <- fig24$simulated[fig24$endpoint == "Insulin (mIU/L)" & fig24$trel == -0.51]
stopifnot(
  # Guard the scalar lookups so an empty match fails loudly instead of making
  # the baseline assertions below vacuously true.
  length(base_glp) == 1L, length(base_ins) == 1L,
  nrow(fig24) == 8L, all(is.finite(fig24$ratio)),
  all(fig24$ratio > 0.45), all(fig24$ratio < 2.2),
  # baselines are the cleanly determined part and must be close
  abs(base_glp - 2.7) < 1.0,
  abs(base_ins - 22.0) < 9
)
```

```{r vpc, fig.height = 5.5}
# Stochastic VPC across all four arms, matching the published figure style.
# Cohort size is well under the 200-per-arm cap.
rxode2::rxSetSeed(20260903)
n_per_arm <- 150L
vpc_ev <- dplyr::bind_rows(lapply(seq_along(ARMS), function(i) {
  x <- as.data.frame(build_arm(ARMS[[i]], 1L))
  x$arm <- names(ARMS)[i]
  x$id <- NULL
  x
}))

vpc <- dplyr::bind_rows(lapply(names(ARMS), function(a) {
  e <- dplyr::filter(vpc_ev, arm == a) |> dplyr::select(-arm)
  s <- rxode2::rxSolve(mod, events = e, nSub = n_per_arm,
                       returnType = "data.frame", useLinCmt = FALSE)
  # rxSolve names the subject column `sim.id` when the cohort is generated by
  # nSub from a single-subject event table, and `id` otherwise.
  s$id <- if (!is.null(s$id)) s$id else s$sim.id
  stopifnot(!is.null(s$id), dplyr::n_distinct(s$id) == n_per_arm)
  s$arm <- a
  s
})) |>
  dplyr::mutate(trel = time - 24 * (N_DAY - 1L),
                arm = factor(arm, levels = names(ARMS)))

vpc_fail <- 100 * mean(is.na(vpc$Cglc))

vpc |>
  dplyr::group_by(arm, trel) |>
  dplyr::summarise(
    Q10 = quantile(Cglc, 0.10, na.rm = TRUE),
    Q50 = quantile(Cglc, 0.50, na.rm = TRUE),
    Q90 = quantile(Cglc, 0.90, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(trel, Q50)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.2) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~arm) +
  coord_cartesian(ylim = c(0, 500)) +
  labs(x = "Time after morning dose (h)", y = "Glucose (mg/dL)",
       title = "Figure 3 style VPC: median and 80% prediction interval",
       caption = "Replicates the layout of Figure 3 of Landersdorfer 2012.")
```

Unsolvable fraction for this cohort: `r sprintf("%.1f%%", vpc_fail)` of
simulated glucose records. The published between-subject variability is very
large on several PD parameters (188% on `S2`, 150% on `S5`, 135% on `S3`), so
extreme log-normal draws can put the glucose-insulin feedback loop into a
numerically stiff regime; the paper reports the same structure without noting
how S-ADAPT handled it. Which draws land there depends on the rxode2 build, so
the fraction is reported rather than asserted and the percentiles below use
`na.rm`.

# Replicating Figure 7 (model subparts)

Figure 7 plots four internal quantities directly, with published axis ranges.
None of these was used to derive `V_glc` or the `S5` prefix.

```{r figure-7, fig.height = 5.5}
fig7 <- sim |>
  dplyr::mutate(
    st_ins    = (0.584 / 1000) * (1 + 1.90 * (Cglp - 1.68)),
    st_glc    = 0.0185 * (1 + 0.0701 * (Cglp - 1.68)),
    ins_secr  = 9.75 * 14.0 * (1 + st_glc * (Cglc - 133)),
    glp_elim  = (DPP4 / 2.80) * 0.641
  )

fig7 |>
  dplyr::select(trel, arm, `A: GLP-1 elimination by DPP-4 (1/h)` = glp_elim,
                `B: ST_ins (L/mIU)` = st_ins, `C: ST_glc (dL/mg)` = st_glc,
                `D: insulin secretion (mIU/L/h)` = ins_secr) |>
  tidyr::pivot_longer(-c(trel, arm), names_to = "panel", values_to = "value") |>
  ggplot(aes(trel, value, colour = arm)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~panel, scales = "free_y") +
  labs(x = "Time after morning dose (h)", y = NULL, colour = NULL,
       title = "Figure 7 panels A-D",
       caption = "Replicates Figure 7 of Landersdorfer 2012.")
```

```{r figure-7-table}
pb7 <- dplyr::filter(fig7, arm == "Placebo")
hi7 <- dplyr::filter(fig7, arm == "100 mg")

fig7_tab <- tibble::tribble(
  ~quantity,                                   ~published,                            ~simulated,
  "7A placebo GLP-1 elimination by DPP-4 (1/h), constant",
    "3.0 (flat line)",                         sprintf("%.2f", 5.0 * 0.641),
  "7A 100 mg GLP-1 elimination by DPP-4 (1/h), range",
    "about 0.05 to 0.25",                      sprintf("%.2f to %.2f", min(hi7$glp_elim), max(hi7$glp_elim)),
  "7B ST_ins (L/mIU), placebo fasting value",
    "about 0.001 (axis 0 to 0.009)",           sprintf("%.5f", min(pb7$st_ins)),
  "7C ST_glc (dL/mg), placebo fasting value",
    "about 0.019 (axis 0.015 to 0.029)",       sprintf("%.4f", min(pb7$st_glc)),
  "7D insulin secretion (mIU/L/h), placebo fasting value",
    "about 180 (axis 150 to 450)",             sprintf("%.0f", min(pb7$ins_secr)),
  "7D insulin secretion (mIU/L/h), placebo peak",
    "about 360",                               sprintf("%.0f", max(pb7$ins_secr))
)

fig7_tab |>
  dplyr::rename("Quantity" = quantity, "Figure 7" = published, "Simulated" = simulated) |>
  knitr::kable(caption = "Figure 7 model subparts vs the packaged model (typical values).")
```

```{r figure-7-gate}
# All deterministic. The fasting values are the cleanly determined part of each
# panel and are asserted tightly against the published axes; the postprandial
# peaks are amplified by the glucose excursion and are recorded as a deviation
# rather than gated (see Assumptions and deviations).
stopifnot(
  abs(5.0 * 0.641 - 3.0) < 0.3,                        # 7A placebo plateau
  max(hi7$glp_elim) < 0.35,                            # 7A 100 mg stays low
  min(pb7$st_ins) > 5e-4, min(pb7$st_ins) < 1.5e-3,    # 7B fasting sits on the axis
  max(pb7$st_ins) < 0.009,                             # 7B stays within the published axis
  min(pb7$st_glc) > 0.018, min(pb7$st_glc) < 0.022,    # 7C fasting
  max(pb7$st_glc) < 0.029,                             # 7C within the published axis
  min(pb7$ins_secr) > 150, min(pb7$ins_secr) < 260     # 7D fasting
)
```

# Published quantitative claims

```{r claims}
th <- function(nm) exp(ui$theta[[nm]])

claims <- tibble::tribble(
  ~claim, ~published, ~reproduced,
  "S2 implies a 35% rise in insulin secretion per 5 pmol/L rise in GLP-1 (Discussion, p. 385)",
    35,   100 * ((1 + th("ls2") * 5) - 1),
  "Complete DPP-4 inhibition raises the active GLP-1 half-life about 2.5-fold (Results, p. 377)",
    2.5,  (th("lkout_glp_lin") + th("lrmaxc") * th("lcf2")) / th("lkout_glp_lin"),
  "Half-life of vildagliptin dissociation from DPP-4 = 1.1 h (companion Abstract)",
    1.1,  log(2) / th("lkoff"),
  "Half-life of vildagliptin hydrolysis by DPP-4 = 6.3 h (companion Abstract)",
    6.3,  log(2) / th("lkdeg"),
  "Endogenous glucose production is physiological (mg/kg/min, 91 kg) -- consistency check on the back-solved V_glc",
    2.0,  th("lbl_glc") * th("lkout_glc") * th("lv_glc") / 60 / 91
) |>
  dplyr::mutate(pct_diff = 100 * (reproduced - published) / published)

claims |>
  dplyr::mutate(dplyr::across(c(published, reproduced, pct_diff), \(x) round(x, 3))) |>
  dplyr::rename("Published claim" = claim, "Published" = published,
                "From the packaged model" = reproduced, "Difference (%)" = pct_diff) |>
  knitr::kable(caption = "Quantitative claims in the source, recomputed from the packaged model.")
```

```{r claims-gate}
# Deterministic algebra on the packaged thetas, so tight bounds are correct.
# The first four are exact restatements of published numbers; the fifth is a
# physiological plausibility band, not a published value.
stopifnot(
  abs(claims$reproduced[1] - 35)  < 0.5,   # 35.05% vs 35%
  abs(claims$reproduced[2] - 2.5) < 0.15,  # 2.548 vs "approximately 2.5"
  abs(claims$reproduced[3] - 1.1) < 0.1,   # 1.133 vs 1.1 h
  abs(claims$reproduced[4] - 6.3) < 0.1,   # 6.301 vs 6.3 h
  claims$reproduced[5] > 1.4, claims$reproduced[5] < 3.0
)
```

# Direction and size of the vildagliptin effect

The paper summarises Figure 6 in words: GLP-1 AUC rises with vildagliptin
exposure, glucose AUC falls, and insulin AUC is "similar among all treatments"
(Results, p. 377). Those are reproduced here as day-5 AUC ratios against
placebo.

```{r auc-ratios}
auc_day <- sim |>
  dplyr::filter(trel >= 0, trel <= 24) |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    GLP1    = PKNCA::pk.calc.auc(Cglp, trel, interval = c(0, 24), method = "linear"),
    Glucose = PKNCA::pk.calc.auc(Cglc, trel, interval = c(0, 24), method = "linear"),
    Insulin = PKNCA::pk.calc.auc(Cins, trel, interval = c(0, 24), method = "linear"),
    .groups = "drop"
  )

ratios <- auc_day |>
  dplyr::mutate(dplyr::across(c(GLP1, Glucose, Insulin),
                              \(x) x / x[arm == "Placebo"])) |>
  dplyr::filter(arm != "Placebo")

ratios |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 3))) |>
  dplyr::rename("Arm" = arm, "GLP-1 AUC ratio" = GLP1,
                "Glucose AUC ratio" = Glucose, "Insulin AUC ratio" = Insulin) |>
  knitr::kable(caption = "Day-5 AUC(0-24 h) ratio vs placebo (typical values). Figure 6 of the source shows individual ratios of about 1-3 for GLP-1, 0.6-1.3 for glucose and 0.6-2.1 for insulin.")
```

```{r auc-ratio-gate}
# Deterministic (zeroRe). Directions are the paper's own qualitative claims and
# the magnitudes are bounded by the observed spread of Figure 6.
stopifnot(
  all(ratios$GLP1 > 1),                                  # GLP-1 AUC rises
  ratios$GLP1[ratios$arm == "100 mg"] > ratios$GLP1[ratios$arm == "10 mg"],
  all(ratios$GLP1 < 3.5),                                # within the Figure 6 range
  all(ratios$Glucose < 1),                               # glucose AUC falls
  ratios$Glucose[ratios$arm == "100 mg"] < ratios$Glucose[ratios$arm == "10 mg"],
  all(ratios$Glucose > 0.55),
  all(abs(ratios$Insulin - 1) < 0.35)                    # insulin "similar among treatments"
)
```

# PKNCA validation (vildagliptin)

Vildagliptin plasma concentration is the one endpoint with conventional PK, so
NCA is run on it over the day-5 morning dosing interval (0 to 10.5 h). The
source paper reports no NCA table; Figure 6 plots individual `AUC_vildagliptin`
values spanning roughly 50-2500 ng/mL*h across the three dose levels, which is
the available published anchor.

```{r pknca}
sim_nca <- vpc |>
  dplyr::filter(arm != "Placebo", trel >= 0, trel <= 10.5) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::transmute(id = paste(arm, id, sep = "-"), arm, time = trel, Cc)

# Guarantee a time-zero record per (arm, id).
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(arm, id, time)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(sim_nca), Cc ~ time | arm + id)

dose_df <- vpc |>
  dplyr::filter(arm != "Placebo") |>
  dplyr::distinct(arm, id) |>
  dplyr::transmute(id = paste(arm, id, sep = "-"), arm, time = 0,
                   amt = as.numeric(ARMS[as.character(arm)]))

dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | arm + id)

intervals <- data.frame(start = 0, end = 10.5,
                        cmax = TRUE, tmax = TRUE, auclast = TRUE)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_tab <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast")) |>
  dplyr::group_by(arm, PPTESTCD) |>
  dplyr::summarise(median = stats::median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median)

nca_tab |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 2))) |>
  dplyr::rename("Arm" = arm, "Cmax (ng/mL)" = cmax, "Tmax (h)" = tmax,
                "AUC0-10.5h (ng*h/mL)" = auclast) |>
  knitr::kable(caption = "Median simulated steady-state NCA for vildagliptin over the morning dosing interval of day 5.")
```

```{r pknca-gate}
auc <- setNames(nca_tab$auclast, as.character(nca_tab$arm))
stopifnot(
  # Cohort medians, so bound loosely and on quantities a transcription error
  # would move by a large factor.
  all(auc > 20), all(auc < 4000),                 # inside the Figure 6 x-axis span
  auc[["100 mg"]] > auc[["25 mg"]], auc[["25 mg"]] > auc[["10 mg"]],
  # TMDD makes exposure MORE than dose proportional: saturable binding to a
  # finite DPP-4 pool means a smaller fraction of a large dose is bound.
  auc[["100 mg"]] / auc[["10 mg"]] > 10,
  all(nca_tab$tmax > 0), all(nca_tab$tmax < 6)
)
```

Exposure is more than dose proportional across the 10-fold dose range, which is
the companion paper's headline PK finding: "Due to limited amounts of DPP-4,
vildagliptin concentrations increased more than proportionally with dose."

# Assumptions and deviations

* **`V_glc` is not published.** Back-solved from the Figure 3 placebo median at
  **250 dL**; sensitivity across the unreported meal timings and food-intake
  durations is 246-255 dL. The Figure 3 comparison above is therefore partly
  circular; Figures 2, 4, 6 and 7 are not. Operator ruling 2026-09-02 chose the
  back-solve over author correspondence.
* **`S5` is applied as 0.584 mL/mIU, not the printed 0.584 L/mIU.** Established
  from Figure 7B's own axis, which plots `ST_ins` over 0 to 0.009 L/mIU and
  equals `S5` at baseline GLP-1. The printed value is kept verbatim in `ini()`
  and converted on one commented line in `model()`.
* **`S1` is used in g^-1 per Table 1, not mg^-1 per the body text**, so gut
  amounts are carried in grams.
* **`R_maxP` is encoded as 13000 nmol (13 umol)**, not the companion table's
  printed "13.0 mmol"; established from that paper's own ">2000-fold"
  tissue:plasma statement.
* **Meal timetable.** Breakfast at 0.5 h and dinner at 11.0 h come from the
  stated design (30 min after each dose, evening dose at 10.5 h). Lunch (5.6 h)
  and the snack (14.5 h) are read from the meal-driven excursions in Figures 2,
  3, 4 and 7. `tk0`, "the actual recorded duration of food intake", is never
  reported and is assumed to be 45 min.
* **Between-subject variability is diagonal only.** Both papers state that a
  full variance-covariance matrix was estimated for the PK and (separately) the
  PD parameters, but only the diagonal is published as BSV (%). Off-diagonals
  are omitted. Between-occasion variability was not included by the authors
  either.
* **Steady state is approached in 5 simulated days, not 28.** Day-4 to day-5
  glucose differs by less than 1% at matched clock times.
* **Known deviation, not gated: the postprandial peaks of the Figure 7D
  insulin-secretion panel overshoot.** The packaged model's placebo insulin
  secretion peaks near 570 mIU/L/h where Figure 7D shows about 360, because the
  typical-value glucose excursion is somewhat sharper than the paper's. The
  fasting values of all four Figure 7 panels agree, and Figure 3's glucose
  median agrees to within a few percent, so the discrepancy is in the peak
  shape rather than in the parameter scale. It most plausibly reflects the
  unreported `tk0` and meal times. This is recorded rather than tuned away.
* **Some stochastic draws may not be solvable.** The published BSV reaches 188%
  on `S2` and 150% on `S5`; extreme log-normal draws drive the glucose-insulin
  feedback into a stiff regime. Which subjects are drawn is not reproducible
  across rxode2 builds, so the VPC above measures and reports the affected
  fraction (0% for the packaged seed and cohort size on the build used here)
  and uses `na.rm` percentiles rather than asserting on it.
* Original observed data are not publicly available; every figure here is a
  simulation from the packaged model compared against digitised or transcribed
  published summaries.
