---
title: "Cefazolin (Lanoiselee 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cefazolin (Lanoiselee 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Lanoiselee_2021_cefazolin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Lanoiselee_2021_cefazolin"))$description`
* Article: <https://doi.org/10.1038/s41598-021-99162-7> (open access, Sci Rep 2021;11:19763)

No supplementary material and no correction notice were found for this article
(EuropePMC, checked 2026-09-29).

# Population

Lanoiselee et al. analysed residual plasma from 100 adults undergoing
cementless primary unilateral total hip arthroplasty at the University Hospital
of Saint-Etienne, France, enrolled in the PORTO tranexamic-acid trial
(NCT02252497) between April 2014 and December 2015. Mean age was 67 years
(range 24-91), mean total body weight 76 kg (range 48-123), 49 were female and
29 were obese (BMI > 30 kg/m^2; 5 above 35 kg/m^2). Mean CKD-EPI creatinine
clearance was 83 mL/min/1.73 m^2 (range 17-129); 8 patients were between 30 and
60 and 1 below 30 mL/min/1.73 m^2 (Table 1). Cefazolin was given as a direct
intravenous bolus at anaesthesia induction: 2000 mg in 96 patients, 3000 mg in
1 and 4000 mg in 3 (dose doubled when BMI > 35 kg/m^2 and body weight > 100 kg).
Samples were drawn 3 and 20 min after the bolus, at the end of surgery, and at
3 h and 8 h, giving 484 total-plasma concentrations (LC-MS/MS, LLOQ 5 mg/L).

The same information is available programmatically via
`readModelDb("Lanoiselee_2021_cefazolin")()$population`.

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Structure: two-compartment, IV bolus into central | n/a | Results, 'Population PK model' |
| `cl = exp(lcl + etalcl) * (CRCL / 80)^e_crcl_cl` | n/a | Methods covariate equation; Table 2 row 'CL (L/h) = theta1 x (CrCL/80)^theta2' |
| `lcl` | log(2.86) L/h | Table 2, Model 2, theta1 (RSE 3.26%) |
| `e_crcl_cl` | 0.79 | Table 2, Model 2, theta2 (RSE 10.6%) |
| `lvc` | log(5.2) L | Table 2, Model 2, Vc (RSE 6.89%) |
| `lq` | log(10.9) L/h | Table 2, Model 2, Q (RSE 11.9%) |
| `lvp` | log(4.56) L | Table 2, Model 2, Vp (RSE 3.07%) |
| `etalcl` | 0.32^2 = 0.1024 | Table 2, Model 2, Omega CL = 32 |
| `etalvc` | 0.57^2 = 0.3249 | Table 2, Model 2, Omega Vc = 57 |
| cov(`etalcl`, `etalvc`) | 0.83 x 0.32 x 0.57 = 0.151392 | Table 2, Model 2, correlation CL-Vc = 0.83 |
| `etalq` | 0.66^2 = 0.4356 | Table 2, Model 2, Omega Q = 66 |
| `etalvp` | 0.10^2 = 0.01 | Table 2, Model 2, Omega Vp = 10 |
| `propSd` | 0.12 | Table 2, Model 2, proportional residual = 12; Methods error-model equation with a = 0 |
| `CRCL` centring value | 80 mL/min/1.73 m^2 | Methods: 'CrCLi is centred on 80' |

# Omega scale

Table 2 prints the random-effect rows as whole numbers (Omega CL = 32, Omega Vc
= 57, ...) and its footnote calls them variances. Monolix reports each
`omega` as the standard deviation of the log-normal random effect, so the
printed numbers could be either 100 x SD or 100 x variance. The two readings
give very different spreads (SD 0.32 versus SD 0.57 on CL), and the paper's own
Table 3 separates them: it reports the probability that total cefazolin stays
above 20 mg/L at 2.01 h and at 4 h after a 2000 mg bolus for four CrCL values.
The chunk below reproduces Table 3 under both readings.

The simulation is deterministic: the random effects are drawn with base R and
passed to the model as data (`zeroRe()` removes the model's own random
effects), and the proportional residual error is integrated analytically with
`pnorm()`. Base R's generator is stable across builds and thread counts, so the
numbers below are reproducible on every machine.

```{r pta}
mod <- readModelDb("Lanoiselee_2021_cefazolin")
ui <- rxode2::rxode(mod)
omega_sd <- ui$omega # the packaged (SD-reading) covariance matrix
eta_names <- colnames(omega_sd)
prop_sd <- ui$theta[["propSd"]]

# Alternative reading: the printed numbers /100 are VARIANCES, with the same
# correlation structure.
omega_var <- local({
  d <- diag(sqrt(sqrt(diag(omega_sd))))
  d %*% stats::cov2cor(omega_sd) %*% d
})
dimnames(omega_var) <- dimnames(omega_sd)

set.seed(20211006)
n_pta <- 200L
z <- matrix(rnorm(n_pta * length(eta_names)), n_pta)
crcl_levels <- c(120, 90, 60, 30)

pta_events <- function(omega, reading) {
  etas <- z %*% chol(omega)
  colnames(etas) <- eta_names
  base <- dplyr::bind_cols(tibble::tibble(sid = seq_len(n_pta)), tibble::as_tibble(etas))
  cohort <- tidyr::crossing(base, CRCL = crcl_levels) |>
    dplyr::mutate(id = dplyr::row_number(), reading = reading)
  dplyr::bind_rows(
    cohort |> dplyr::mutate(time = 0, evid = 1L, amt = 2000, cmt = "central"),
    cohort |> dplyr::mutate(time = 2.01, evid = 0L, amt = 0, cmt = "central"),
    cohort |> dplyr::mutate(time = 4, evid = 0L, amt = 0, cmt = "central")
  ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

solve_with_data_etas <- function(ev) {
  withCallingHandlers(
    rxode2::rxSolve(rxode2::zeroRe(mod), events = ev, keep = c("CRCL", "reading")),
    warning = function(w) {
      if (grepl("without 'omega'", conditionMessage(w), fixed = TRUE)) {
        invokeRestart("muffleWarning")
      }
    }
  )
}

pta_sim <- dplyr::bind_rows(
  as.data.frame(solve_with_data_etas(pta_events(omega_sd, "SD (packaged)"))),
  as.data.frame(solve_with_data_etas(pta_events(omega_var, "variance")))
)
stopifnot(nrow(pta_sim) == 2L * 2L * n_pta * length(crcl_levels))

pta_tab <- pta_sim |>
  dplyr::group_by(reading, CRCL, time) |>
  dplyr::summarise(
    pta_ipred = 100 * mean(Cc > 20),
    pta_obs = 100 * mean(stats::pnorm(20, Cc, prop_sd * Cc, lower.tail = FALSE)),
    .groups = "drop"
  )

published_pta <- tibble::tribble(
  ~CRCL, ~time, ~pta_published,
  120, 2.01, 100,
  90, 2.01, 100,
  60, 2.01, 100,
  30, 2.01, 100,
  120, 4, 94.5,
  90, 4, 99.8,
  60, 4, 100,
  30, 4, 100
)

pta_cmp <- dplyr::inner_join(pta_tab, published_pta, by = c("CRCL", "time"))
stopifnot(nrow(pta_cmp) == 16L)

pta_cmp |>
  dplyr::arrange(reading, dplyr::desc(CRCL), time) |>
  dplyr::mutate(time = sprintf("%.2f", time)) |>
  dplyr::rename(
    "Omega reading" = reading,
    "CrCL (mL/min/1.73 m^2)" = CRCL,
    "Time (h)" = time,
    "PTA, no residual (%)" = pta_ipred,
    "PTA, with residual (%)" = pta_obs,
    "Published PTA (%)" = pta_published
  ) |>
  knitr::kable(digits = 1, caption = "Replicates Table 3 of Lanoiselee 2021: probability of total cefazolin > 20 mg/L after a 2000 mg bolus.")

pta_cell <- function(rd, crcl, t) {
  v <- pta_cmp$pta_obs[pta_cmp$reading == rd & pta_cmp$CRCL == crcl & pta_cmp$time == t]
  if (length(v) != 1L) stop("no unique PTA row for ", rd, " / ", crcl, " / ", t)
  v
}
stopifnot(
  # SD reading reproduces every Table 3 cell within 3 percentage points.
  all(abs(pta_cmp$pta_obs - pta_cmp$pta_published)[pta_cmp$reading == "SD (packaged)"] < 3),
  # Mutation control: the variance reading misses the CrCL 120 and 90 cells
  # at 4 h by far more than that, so this table does discriminate the scale.
  pta_cell("variance", 120, 4) < 94.5 - 8,
  pta_cell("variance", 90, 4) < 99.8 - 4
)
```

Under the SD reading every Table 3 cell is reproduced within 3 percentage
points (published 94.5% and 99.8% at 4 h for CrCL 120 and 90 against
`r round(pta_cell("SD (packaged)", 120, 4), 1)`% and
`r round(pta_cell("SD (packaged)", 90, 4), 1)`% here; 100% elsewhere). Under the
variance reading the 4 h values at CrCL 120 and 90 fall to
`r round(pta_cell("variance", 120, 4), 1)`% and
`r round(pta_cell("variance", 90, 4), 1)`%.
The packaged model therefore uses the printed values as 100 x SD, and the
proportional residual error (printed 12) is read on the same scale as
`propSd = 0.12`.

# Typical-value checks

A deterministic typical-value solve at the four CrCL values of Figure 4 is
compared with the closed-form two-compartment bolus solution.

```{r typical}
t_grid <- sort(unique(c(seq(0, 1, by = 0.05), seq(1, 24, by = 0.25), 2.01)))
typ_events <- tidyr::crossing(id = seq_along(crcl_levels), time = t_grid) |>
  dplyr::mutate(evid = 0L, amt = 0) |>
  dplyr::bind_rows(tibble::tibble(id = seq_along(crcl_levels), time = 0, evid = 1L, amt = 2000)) |>
  dplyr::mutate(cmt = "central", CRCL = crcl_levels[id]) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

typ <- rxode2::rxSolve(rxode2::zeroRe(mod), events = typ_events, keep = "CRCL",
                       rtol = 1e-10, atol = 1e-12) |>
  as.data.frame()

th <- ui$theta
closed_form <- function(t, dose, cl, vc, q, vp) {
  k10 <- cl / vc; k12 <- q / vc; k21 <- q / vp
  s <- k10 + k12 + k21
  r <- sqrt(s^2 - 4 * k10 * k21)
  a <- (s + r) / 2; b <- (s - r) / 2
  dose / vc * ((a - k21) / (a - b) * exp(-a * t) + (k21 - b) / (a - b) * exp(-b * t))
}
typ <- typ |>
  dplyr::mutate(
    cl_typ = exp(th[["lcl"]]) * (CRCL / 80)^th[["e_crcl_cl"]],
    Cc_cf = closed_form(time, 2000, cl_typ, exp(th[["lvc"]]), exp(th[["lq"]]), exp(th[["lvp"]]))
  )
rel_err <- max(abs(typ$Cc / typ$Cc_cf - 1))
stopifnot(rel_err < 1e-6)
rel_err
```

The ODE solution matches the closed form to a maximum relative error of
`r signif(rel_err, 2)`.

# Replicate Figure 4

Figure 4 shows the concentration-time course after a 2000 mg bolus at CrCL 30,
60, 90 and 120 mL/min/1.73 m^2, with the typical profile, an interpatient
variability band and the 20 and 360 mg/L thresholds. The stochastic cohort
below uses 200 virtual patients per CrCL level.

```{r cohort}
rxode2::rxSetSeed(20211006)
n_sub <- 200L
vpc_times <- sort(unique(c(0, 3 / 60, 20 / 60, seq(0.25, 8, by = 0.25))))
vpc_events <- tidyr::crossing(sid = seq_len(n_sub), CRCL = crcl_levels) |>
  dplyr::mutate(id = dplyr::row_number())
vpc_events <- dplyr::bind_rows(
  vpc_events |> dplyr::mutate(time = 0, evid = 1L, amt = 2000),
  tidyr::crossing(vpc_events, time = vpc_times) |> dplyr::mutate(evid = 0L, amt = 0)
) |>
  dplyr::mutate(cmt = "central") |>
  dplyr::arrange(id, time, dplyr::desc(evid))
stopifnot(dplyr::n_distinct(vpc_events$id) == n_sub * length(crcl_levels))

vpc <- rxode2::rxSolve(mod, events = vpc_events, keep = "CRCL") |> as.data.frame()
```

```{r figure-4, fig.width = 8, fig.height = 3.5}
band <- vpc |>
  dplyr::filter(time > 0) |>
  dplyr::group_by(CRCL, time) |>
  dplyr::summarise(
    lo = quantile(ipredSim, 0.05), hi = quantile(ipredSim, 0.95),
    .groups = "drop"
  )
typ_plot <- typ |> dplyr::filter(time > 0, time <= 8)

ggplot() +
  geom_ribbon(data = band, aes(time, ymin = lo, ymax = hi), fill = "steelblue", alpha = 0.3) +
  geom_line(data = typ_plot, aes(time, Cc)) +
  geom_hline(yintercept = c(20, 360), linetype = "dashed", colour = "red") +
  facet_wrap(~ paste("CrCL", CRCL, "mL/min/1.73 m^2"), nrow = 1) +
  scale_y_log10() +
  labs(
    x = "Time after bolus (h)", y = "Total cefazolin (mg/L)",
    caption = "Replicates Figure 4 of Lanoiselee 2021. Line: typical value; band: 5th-95th percentile of individual predictions."
  )
```

The paper states that after a 2000 mg bolus concentrations stay above 20 mg/L
throughout surgery regardless of renal function. The typical-value
concentration at the mean time of skin closure (2.01 h) and at 4 h is shown
below; the lowest value, at CrCL 120, is still above the threshold at 4 h.

```{r typical-thresholds}
typ_at <- typ |>
  dplyr::filter(abs(time - 0.05) < 1e-9 | abs(time - 2.01) < 1e-9 | abs(time - 4) < 1e-9) |>
  dplyr::mutate(time = round(time, 2)) |>
  dplyr::select(CRCL, time, Cc) |>
  tidyr::pivot_wider(names_from = time, values_from = Cc, names_prefix = "t_")
stopifnot(nrow(typ_at) == 4L, all(typ_at$t_4 > 20))
typ_at |>
  dplyr::rename(
    "CrCL (mL/min/1.73 m^2)" = CRCL,
    "Cc at 3 min (mg/L)" = t_0.05,
    "Cc at 2.01 h (mg/L)" = t_2.01,
    "Cc at 4 h (mg/L)" = t_4
  ) |>
  knitr::kable(digits = 1, caption = "Typical total cefazolin concentrations after a 2000 mg bolus.")
```

# Figure 5 regimens (illustrative)

Figure 5 compares a 4000 mg bolus followed by 2000 mg at 4 h with a 2000 mg
bolus followed by 1000 mg at 4 h in the five patients with BMI > 35 kg/m^2 and
body weight > 100 kg. Those curves use the patients' individual (post hoc)
parameters, which are not published, so they cannot be reproduced. The chunk
below shows the same two regimens for a typical patient at the centring CrCL
of 80 mL/min/1.73 m^2. Because the model is linear, the higher regimen is
exactly twice the lower one.

```{r figure-5, fig.width = 6, fig.height = 3.5}
reg_events <- dplyr::bind_rows(
  tibble::tibble(id = 1L, regimen = "2000 mg + 1000 mg at 4 h", time = c(0, 4), amt = c(2000, 1000)),
  tibble::tibble(id = 2L, regimen = "4000 mg + 2000 mg at 4 h", time = c(0, 4), amt = c(4000, 2000))
) |>
  dplyr::mutate(evid = 1L)
reg_events <- dplyr::bind_rows(
  reg_events,
  tidyr::crossing(dplyr::distinct(reg_events, id, regimen), time = seq(0.05, 8, by = 0.05)) |>
    dplyr::mutate(evid = 0L, amt = 0)
) |>
  dplyr::mutate(cmt = "central", CRCL = 80) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

reg <- rxode2::rxSolve(rxode2::zeroRe(mod), events = reg_events, keep = "regimen",
                       rtol = 1e-10, atol = 1e-12) |>
  as.data.frame()
ratio <- reg$Cc[reg$id == 2] / reg$Cc[reg$id == 1]
stopifnot(length(ratio) > 0, max(abs(ratio / 2 - 1)) < 1e-6)

ggplot(reg, aes(time, Cc, colour = regimen)) +
  geom_line() +
  geom_hline(yintercept = c(20, 360), linetype = "dashed", colour = "red") +
  scale_y_log10() +
  labs(x = "Time after first bolus (h)", y = "Total cefazolin (mg/L)", colour = NULL,
       caption = "Typical patient, CrCL 80 mL/min/1.73 m^2 (cf. Figure 5 of Lanoiselee 2021).") +
  theme(legend.position = "bottom")
```

# PKNCA validation

The paper reports no NCA table. PKNCA is run on the stochastic Figure 4 cohort,
extended to 48 h so the terminal phase is captured, and each subject's
`AUC0-inf` is checked against `Dose / CL_i` using that subject's own simulated
clearance.

```{r pknca}
nca_times <- sort(unique(c(0, 3 / 60, 20 / 60, seq(0.25, 2, by = 0.25), seq(2.5, 12, by = 0.5), seq(13, 48, by = 1))))
nca_ev <- tidyr::crossing(sid = seq_len(n_sub), CRCL = crcl_levels) |>
  dplyr::mutate(id = dplyr::row_number(), treatment = paste0("CrCL ", CRCL))
nca_ev <- dplyr::bind_rows(
  nca_ev |> dplyr::mutate(time = 0, evid = 1L, amt = 2000),
  tidyr::crossing(nca_ev, time = nca_times) |> dplyr::mutate(evid = 0L, amt = 0)
) |>
  dplyr::mutate(cmt = "central") |>
  dplyr::arrange(id, time, dplyr::desc(evid))

nca_sim <- rxode2::rxSolve(mod, events = nca_ev, keep = c("CRCL", "treatment")) |>
  as.data.frame()
stopifnot(all(nca_sim$ipredSim >= -1e-6 * max(nca_sim$ipredSim)))

conc_df <- nca_sim |>
  dplyr::filter(!is.na(ipredSim)) |>
  dplyr::transmute(id, time, treatment, Cc = pmax(ipredSim, 0))
dose_df <- nca_ev |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)
intervals <- data.frame(start = 0, end = Inf, cmax = TRUE, aucinf.obs = TRUE, half.life = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_ind <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "aucinf.obs", "half.life")) |>
  dplyr::select(id, treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::left_join(dplyr::distinct(nca_sim, id, cl, vc), by = "id") |>
  dplyr::mutate(
    pct_diff = 100 * (aucinf.obs / (2000 / cl) - 1),
    cmax_diff = 100 * (cmax / (2000 / vc) - 1)
  )
stopifnot(nrow(nca_ind) == n_sub * length(crcl_levels))
stopifnot(
  abs(median(nca_ind$pct_diff)) < 1,
  quantile(abs(nca_ind$pct_diff), 0.9) < 3,
  # Time-zero record follows the bolus, so Cmax is the instantaneous Dose/Vc.
  max(abs(nca_ind$cmax_diff)) < 1e-3
)

nca_ind |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    cmax = median(cmax), aucinf = median(aucinf.obs), t_half = median(half.life),
    auc_vs_dose_cl = median(pct_diff), .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(as.numeric(sub("CrCL ", "", treatment)))) |>
  dplyr::rename(
    "Group" = treatment,
    "Median Cmax (mg/L)" = cmax,
    "Median AUC0-inf (mg*h/L)" = aucinf,
    "Median terminal t1/2 (h)" = t_half,
    "Median AUC0-inf vs Dose/CL (%)" = auc_vs_dose_cl
  ) |>
  knitr::kable(digits = 1, caption = "PKNCA summary of the simulated 2000 mg bolus cohort (200 patients per CrCL level).")
```

`AUC0-inf` agrees with `Dose / CL_i` subject by subject (median difference
below 1%), confirming dose, clearance and units are consistent. Because the
time-zero record is taken after the bolus, PKNCA's Cmax equals each subject's
instantaneous `Dose / Vc`.

# Assumptions and deviations

- **Omega and residual-error scale.** Table 2 labels the random-effect rows as
  variances, but the printed numbers are 100 x the Monolix log-scale SD. The
  maintainers settled this against the paper's own Table 3 PTA values (see
  "Omega scale" above). The proportional residual error printed as 12 is taken
  as `propSd = 0.12` on the same basis; Table 3 is not sensitive enough to
  test it separately.
- **CrCL units.** Table 1, Table 3 and the Figure 4 caption give CKD-EPI CrCL in
  mL/min/1.73 m^2, the native CKD-EPI unit, while the Table 2 and Table 3
  footnotes say mL/min. The model uses the BSA-normalized `CRCL` covariate.
- **Residual error in Table 3.** The paper does not say whether its PTA was
  computed on individual predictions or on simulated observations; both are
  shown and differ by at most
  `r round(max(abs(pta_cmp$pta_obs - pta_cmp$pta_ipred)), 1)` percentage points.
- **Figure 5** uses individual post hoc parameters that are not published; the
  regimens are shown for a typical patient only.
- **Covariates tested but not retained** (age, total body weight, BMI, lean
  body weight and sex) are recorded in the model's `covariatesDataExcluded`
  metadata. Cockcroft-Gault CrCL, tested as an alternative renal-function
  estimator to CKD-EPI, is noted in the `CRCL` covariate entry.
