---
title: "Pediatric sulbactam-durlobactam dose selection (Larson 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pediatric sulbactam-durlobactam dose selection (Larson 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r citation, echo = FALSE, results = "asis"}
cat("* Citation: ",
    readModelDb("Larson_2026_sulbactam_durlobactam_pediatric_allometry")()$reference,
    "\n", sep = "")
```

* Abstract: <https://doi.org/10.1093/ofid/ofaf695.659> (Open Forum Infect Dis
  2026;13(Suppl 1):S405; PMCID `PMC12792777`, open access)
* Poster: `IST_Poster-PPK_SUD_Pediatric_Oct2025_FINAL.pdf`, presented as poster
  P-444 at IDWeek 2025 (Atlanta, GA) and posted publicly by Innoviva Specialty
  Therapeutics. **The poster, not the abstract, is the source of the model
  specification** --- the abstract contains no parameter values, no equations
  and no covariate structure.

Note the **author order differs between the two records**. The indexed abstract
is Larson, O'Donnell, Tanudra, Cammarata, Rubino; the poster is Cammarata,
Larson, Tanudra, O'Donnell, Bhavnani, Rubino. The packaged files are named for
the abstract's first author, which is the citable, PubMed-indexed record.

Sulbactam-durlobactam (XACDURO) is approved in the United States for
hospital-acquired and ventilator-associated bacterial pneumonia caused by
susceptible *Acinetobacter baumannii-calcoaceticus* complex in patients aged 18
years and older. No pediatric clinical trial data existed when this analysis
was done, so the pediatric dosing regimens for the Phase 1b trial
CS2514-2023-002 (NCT06801223) were selected by scaling the published adult
population PK model.

## Two models, one change

The poster explored **two** approaches, and both are packaged:

```{r models}
mod_allo <- readModelDb("Larson_2026_sulbactam_durlobactam_pediatric_allometry")
mod_crcl <- readModelDb("Larson_2026_sulbactam_durlobactam_pediatric_allometry_crcl")
mod_adult <- readModelDb("Cammarata_2024_sulbactam_durlobactam")
```

**Adult with Allometry** (`..._pediatric_allometry`) is the model that guided
dose selection. Poster Methods:

> The first approach utilized the full covariate model from adults with the
> only change being the use of allometric scaling with fixed exponents of 0.75
> for clearance and 1.0 for volume of distribution

with a Figure 1 note adding

> WTKG effects added to Q and Vp for both drugs to account for likely changes
> in these parameters in children

**Allometry + CL~CR~** (`..._pediatric_allometry_crcl`) is the alternative:

> The second approach removed all covariate relationships from the Adult
> Allometry model except for body weight (on all parameters) and renal function
> (on renal clearance only)

The parent adult model is `Cammarata_2024_sulbactam_durlobactam` --- a
four-compartment (two per drug) linear model in which total clearance of each
drug is the sum of a renal arm scaled by baseline BSA-normalized creatinine
clearance and a non-renal arm:

$$
\mathrm{CL_R} = \mathrm{CL}\cdot \mathrm{FE}\cdot\left(\frac{\mathrm{CLcr}}{100}\right)^{\theta_{\mathrm{CLcr}}},
\qquad
\mathrm{CL_{NR}} = \mathrm{CL}\cdot(1-\mathrm{FE})
$$

The pediatric adaptation replaces the adult body-weight terms:

$$
\mathrm{CL_T} = (\mathrm{CL_R}+\mathrm{CL_{NR}})
  \cdot\left(\frac{\mathrm{WT}}{75}\right)^{0.75}\cdots,
\qquad
\mathrm{V_c},\mathrm{V_p} \propto \left(\frac{\mathrm{WT}}{75}\right)^{1.0},
\qquad
\mathrm{Q} \propto \left(\frac{\mathrm{WT}}{75}\right)^{0.75}
$$

# Population

This is a **simulation** population, not an estimation data set. Every parameter
value was estimated in adults (373 subjects, 5,188 plasma concentrations, six
Phase 1 + one Phase 2 + one Phase 3 study; Cammarata 2024) and re-used
unchanged. The poster generated 8,000 hypothetical pediatric patients, 1,000
per cohort, spanning birth (28 weeks of gestation) to < 18 years, with age- and
sex-appropriate body size from CDC growth charts and normal renal function.

```{r population}
str(readModelDb("Larson_2026_sulbactam_durlobactam_pediatric_allometry")()$population,
    max.level = 1)
```

# Source trace

Almost every `ini()` value in these two models is **inherited unchanged** from
the adult parent model; the poster re-used the adult estimates and changed only
the body-size terms. The table below therefore separates the two provenances.

| Equation / parameter | Value | Source location |
|---|---|---|
| `e_wt_cl`, `e_wt_cl_sbt` | 0.75 (FIXED) | Poster P-444 Methods, first bullet ("fixed exponents of 0.75 for clearance") |
| `e_wt_vc`, `e_wt_vc_sbt` | 1.0 (FIXED) | Poster P-444 Methods, first bullet ("1.0 for volume of distribution") |
| `e_wt_q`, `e_wt_q_sbt` | 0.75 (FIXED) | Poster P-444 Figure 1 notes ("WTKG effects added to Q and Vp for both drugs") + Methods bullet (0.75 for clearance) |
| `e_wt_vp`, `e_wt_vp_sbt` | 1.0 (FIXED) | Poster P-444 Figure 1 notes + Methods bullet (1.0 for volume) |
| Covariate set retained (model 1) | full adult model | Poster P-444 Methods, first bullet ("the full covariate model from adults") |
| Covariate set retained (model 2) | WT + CLcr on renal CL only | Poster P-444 Methods, second bullet |
| Infection type = bacteremia in the simulations | n/a | Poster P-444 Figure 1 notes ("Infection type was set to 'Bacteremia'") |
| Region = non-East-Asian in the simulations | n/a | Poster P-444 Figure 1 notes ("East Asian flag was set to 'not East Asian'") |
| CLcr units mL/min/1.73 m^2 | n/a | Poster P-444 Figure 1 notes ("CL CR represents GFR normalized to BSA") |
| Pediatric dosing regimens | see below | Poster P-444 Table 2 |
| Adult reference exposures | see below | Poster P-444 Table 1 |
| PK/PD targets | 50% fT>MIC (SUL); fAUC/MIC 10 (DUR) | Poster P-444 Methods, PK/PD drivers |
| `lcl`, `lvc`, `lq`, `lvp` (durlobactam) | 9.33 L/h, 12.5 L, 4.43 L/h, 5.83 L | *Inherited*: Cammarata 2024 Table 1, Durlobactam |
| `lcl_sbt`, `lvc_sbt`, `lq_sbt`, `lvp_sbt` | 13.5 L/h, 12 L, 7.88 L/h, 6.99 L | *Inherited*: Cammarata 2024 Table 1, Sulbactam |
| `f_renal`, `f_renal_sbt` | 0.479 (FIXED) | *Inherited*: Cammarata 2024 Table 1, `FE (%)` |
| `e_crcl_cl_renal`, `e_crcl_cl_renal_sbt` | 0.875, 1.14 | *Inherited*: Cammarata 2024 Table 1, `CL R, CLcr power` |
| Infection / region / severe-renal-impairment shifts | see model file | *Inherited*: Cammarata 2024 Table 1 |
| `e_hemodial_active_cl(_sbt)` | log(6.24), log(8.19) | *Inherited*: Cammarata 2024 Table S4 |
| `lrelf`, `lrelf_sbt` | log(0.372), log(0.533) | *Inherited*: Cammarata 2024 Table S6 |
| All IIV and residual-error terms | see model file | *Inherited*: Cammarata 2024 Tables 1 and S4/S6 |
| Reference weight 75 kg | n/a | *Inherited*: Cammarata 2024 Results reference subject; the poster does not restate it (see Errata) |
| Reference CLcr 100 mL/min/1.73 m^2 | n/a | *Inherited*: Cammarata 2024 Results reference subject |

## Mechanical check that the inheritance is exact

The claim "the only change is the allometric scaling" is machine-checkable:
compare every `ini()` estimate against the adult parent model.

```{r inheritance-check}
ui <- function(m) rxode2::rxode(m)

tv <- function(u) {
  d <- u$iniDf
  d <- d[is.na(d$neta1), ]
  stats::setNames(d$est, d$name)
}

a <- tv(ui(mod_adult))
b <- tv(ui(mod_allo))
d2 <- tv(ui(mod_crcl))

shared     <- intersect(names(a), names(b))
differing  <- shared[abs(a[shared] - b[shared]) > 1e-12]
added      <- setdiff(names(b), names(a))
dropped_1  <- setdiff(names(a), names(b))

tibble::tibble(
  Parameter  = differing,
  Adult      = unname(a[differing]),
  Pediatric  = unname(b[differing])
) |>
  dplyr::rename("Adult (Cammarata 2024)" = Adult,
                "Pediatric (allometry)"  = Pediatric) |>
  knitr::kable(
    caption = paste(
      "Every ini() parameter that differs between the adult parent model and",
      "the 'Adult with Allometry' pediatric model. These four rows are the",
      "entire change the poster describes."
    )
  )
```

```{r inheritance-assert}
# Structural gate. The poster says the ONLY change is the allometric scaling,
# so exactly the four body-weight exponents may differ, the four new Q / Vp
# exponents may be added, and NOTHING may be dropped.
stopifnot(
  setequal(differing, c("e_wt_cl", "e_wt_vc", "e_wt_cl_sbt", "e_wt_vc_sbt")),
  setequal(added, c("e_wt_q", "e_wt_vp", "e_wt_q_sbt", "e_wt_vp_sbt")),
  length(dropped_1) == 0L,
  # The eight allometric exponents must be FIXED, not estimated.
  all(ui(mod_allo)$iniDf$fix[grepl("^e_wt_", ui(mod_allo)$iniDf$name)]),
  all(ui(mod_crcl)$iniDf$fix[grepl("^e_wt_", ui(mod_crcl)$iniDf$name)]),
  # ... and take exactly the poster's two values.
  all(b[c("e_wt_cl", "e_wt_cl_sbt", "e_wt_q", "e_wt_q_sbt")] == 0.75),
  all(b[c("e_wt_vc", "e_wt_vc_sbt", "e_wt_vp", "e_wt_vp_sbt")] == 1.0)
)

# The second approach keeps the same allometry but drops every covariate
# relationship except CLcr on the renal arm.
shared2 <- intersect(names(a), names(d2))
stopifnot(
  setequal(shared2[abs(a[shared2] - d2[shared2]) > 1e-12],
           c("e_wt_cl", "e_wt_vc", "e_wt_cl_sbt", "e_wt_vc_sbt")),
  # CLcr on the renal arm survives ...
  all(c("e_crcl_cl_renal", "e_crcl_cl_renal_sbt") %in% names(d2)),
  # ... and every infection-type, region, severe-renal-impairment and
  # hemodialysis term is gone.
  !any(grepl("^e_(habp|vabp|cuti|bacteremia|ap|region|renalimp|hemodial)",
             names(d2)))
)

# Both models must keep explicit ODEs. A cl / vc pair can make rxode2 silently
# auto-solve a model and discard the d/dt() block; linCmt() would be non-empty.
stopifnot(length(ui(mod_allo)$linCmt) == 0L,
          length(ui(mod_crcl)$linCmt) == 0L)
```

# Structural checks

## The pediatric model reduces exactly to the adult model at 75 kg

Every allometric factor is `(WT / 75)^k`, so at WT = 75 kg all four factors are
exactly 1 and the "Adult with Allometry" model must return the adult model's
own typical-value parameters. This is an exact identity, not an approximation.

```{r solve-helpers}
#' Build a dose + observation event table for the four-compartment SUD model.
#'
#' Doses are 3-hour IV infusions of `dose_mg` of EACH drug into its own central
#' compartment. Observations sit on the durlobactam ODE state `central`;
#' rxode2 returns every algebraic observable (Cc, Cc_sbt, Celf, Celf_sbt) as a
#' column at those rows.
make_events <- function(subj, dose_mg, tau, tmax, by = 0.25) {
  doses <- subj |>
    tidyr::crossing(time = seq(0, tmax - tau, by = tau),
                    cmt = c("central", "central_sbt")) |>
    dplyr::mutate(evid = 1L, amt = dose_mg, rate = dose_mg / 3,
                  dvid = NA_integer_)
  obs <- subj |>
    tidyr::crossing(time = seq(0, tmax, by = by)) |>
    dplyr::mutate(evid = 0L, amt = NA_real_, rate = NA_real_,
                  cmt = "central", dvid = 1L)
  dplyr::bind_rows(doses, obs) |> dplyr::arrange(id, time, dplyr::desc(evid))
}

#' Solve at typical values (no IIV, no residual error).
solve_typical <- function(mod, events, keep) {
  rxode2::rxSolve(
    rxode2::zeroRe(mod), events = events, keep = keep,
    omega = NA, sigma = NA,
    # rxode2's automatic ODE -> linCmt() conversion mis-maps the outputs of
    # this four-compartment two-analyte model; disable it.
    useLinCmt = FALSE
  ) |> as.data.frame()
}

trapz <- function(x, y) sum(diff(x) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)

# The poster's simulation settings, applied to every cohort:
#   infection type = bacteremia, region = not East Asian, normal renal function.
POSTER_COVS <- list(
  CRCL                = 100,   # "normal renal function"; the model's own reference
  RENALIMP_SEV        = 0,
  REGION_EASTASIA     = 0,
  DIS_HABP            = 0, DIS_VABP = 0, DIS_CUTI = 0,
  DIS_BACTEREMIA      = 1, DIS_AP   = 0,
  RRT_HEMODIAL_ACTIVE = 0,
  STUDY_SULDUR_PHASE2 = 0, STUDY_SULDUR_PHASE3 = 1
)
```

```{r adult-recovery}
ref_subj <- tibble::as_tibble(c(list(id = 1L, WT = 75), POSTER_COVS))

ref_ev  <- make_events(ref_subj, dose_mg = 1000, tau = 6, tmax = 12)
sim_ped <- solve_typical(mod_allo,  ref_ev, keep = "WT")
sim_ad  <- solve_typical(mod_adult, ref_ev, keep = "WT")

pars <- c("cl", "vc", "q", "vp", "cl_sbt", "vc_sbt", "q_sbt", "vp_sbt")
cmp_75 <- tibble::tibble(
  Parameter  = pars,
  Adult      = vapply(pars, \(p) sim_ad[[p]][1],  numeric(1)),
  Pediatric  = vapply(pars, \(p) sim_ped[[p]][1], numeric(1))
) |>
  dplyr::mutate(`Ratio` = Pediatric / Adult)

cmp_75 |>
  dplyr::rename("Adult model at 75 kg" = Adult,
                "Pediatric allometry model at 75 kg" = Pediatric) |>
  knitr::kable(
    digits  = 6,
    caption = paste(
      "At the 75 kg reference weight every allometric factor is exactly 1, so",
      "the pediatric model must return the adult model's typical-value",
      "clearances (L/h) and volumes (L) identically."
    )
  )

# Exact identity: this is deterministic algebra, not a simulated statistic, so
# the bound is machine precision rather than a tolerance.
stopifnot(max(abs(cmp_75$Ratio - 1)) < 1e-10)
```

## The allometric exponents are exactly 0.75 and 1.0

Solving at two weights recovers the exponents by
$\log(P_2/P_1)\,/\,\log(\mathrm{WT}_2/\mathrm{WT}_1)$. With every other
covariate held fixed this must return 0.75 for the two clearances and 1.0 for
the two volumes, for both drugs.

```{r exponent-recovery}
w_subj <- tibble::as_tibble(c(list(id = 1:2, WT = c(10, 60)), POSTER_COVS))
w_sim  <- solve_typical(mod_allo, make_events(w_subj, 500, 6, 6), keep = "WT")

first_by_id <- w_sim |> dplyr::group_by(id) |> dplyr::slice(1) |> dplyr::ungroup()

exps <- tibble::tibble(
  Parameter = pars,
  Recovered = vapply(
    pars,
    \(p) log(first_by_id[[p]][2] / first_by_id[[p]][1]) / log(60 / 10),
    numeric(1)
  ),
  Expected  = c(0.75, 1.0, 0.75, 1.0, 0.75, 1.0, 0.75, 1.0)
)

exps |>
  dplyr::rename("Exponent recovered from the solve" = Recovered,
                "Poster P-444 value" = Expected) |>
  knitr::kable(
    digits  = 8,
    caption = paste(
      "Allometric exponents recovered by solving the model at 10 kg and 60 kg.",
      "Clearance terms (cl, q) take 0.75 and volume terms (vc, vp) take 1.0,",
      "for both durlobactam and sulbactam."
    )
  )

stopifnot(max(abs(exps$Recovered - exps$Expected)) < 1e-8)
```

## Mass balance: CL x AUC~tau~ = dose at steady state

For an intravenous drug with no bioavailability term, steady-state exposure
over one dosing interval must equal dose / clearance exactly. This is the
closed-form gate on the whole ODE system --- a mis-transcribed clearance,
volume or infusion rate breaks it.

```{r mass-balance}
mb_subj <- tibble::as_tibble(c(list(id = 1:3, WT = c(4, 20, 60)), POSTER_COVS))

# q6h to 48 h; durlobactam and sulbactam half-lives are only a few hours, so
# 42-48 h is deep in steady state. Fine grid so the trapezoid is accurate.
mb_sim <- solve_typical(mod_allo, make_events(mb_subj, 400, 6, 48, by = 0.02),
                        keep = "WT")

mb <- mb_sim |>
  dplyr::filter(!is.na(Cc), time >= 42, time <= 48) |>
  dplyr::group_by(id, WT) |>
  dplyr::summarise(
    auc_dur = trapz(time, Cc),
    auc_sbt = trapz(time, Cc_sbt),
    cl_dur  = dplyr::first(cl),
    cl_sbt  = dplyr::first(cl_sbt),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    `Durlobactam CL x AUC / dose` = cl_dur * auc_dur / 400,
    `Sulbactam CL x AUC / dose`   = cl_sbt * auc_sbt / 400
  )

mb |>
  dplyr::select(`WT (kg)` = WT, dplyr::starts_with("Durlobactam"),
                dplyr::starts_with("Sulbactam")) |>
  knitr::kable(
    digits  = 4,
    caption = paste(
      "Steady-state mass balance over the 42-48 h dosing interval, 400 mg of",
      "each drug q6h as a 3-hour infusion. The ratio must be 1."
    )
  )

# Residual departure is trapezoidal-integration error on a 0.02 h grid, not
# model error, so the bound is tight.
stopifnot(
  max(abs(mb$`Durlobactam CL x AUC / dose` - 1)) < 0.002,
  max(abs(mb$`Sulbactam CL x AUC / dose`   - 1)) < 0.002
)
```

# The Phase 1b dosing regimens (Table 2)

```{r regimens}
# Poster P-444 Table 2, verbatim. The 1 g cap applies to Cohorts 1 and 2
# (footnote [a]: "Not to exceed a total dose of 1 g SUL and 1 g DUR").
regimens <- tibble::tribble(
  ~cohort,                    ~age_range,       ~mg_per_kg, ~tau, ~cap_mg,
  "Cohort 1",                 "12 to < 18 y",   25,          6,    1000,
  "Cohort 2",                 "6 to < 12 y",    25,          6,    1000,
  "Cohort 3",                 "1 to < 6 y",     25,          6,    Inf,
  "Cohort 4",                 "3 mo to < 1 y",  25,          6,    Inf,
  "Cohort 5 term, 2-<3 mo",   "2 to < 3 mo",    25,          6,    Inf,
  "Cohort 5 preterm, 2-<3 mo","2 to < 3 mo",    20,          8,    Inf,
  "Cohort 5 term, birth-<2 mo","birth to < 2 mo", 25,        8,    Inf,
  "Cohort 5 preterm, birth-<2 mo","birth to < 2 mo", 20,     12,   Inf
)

regimens |>
  dplyr::transmute(
    Cohort = cohort,
    "Age range" = age_range,
    "Regimen" = paste0(mg_per_kg, " mg/kg of each drug q", tau, "h",
                       ifelse(is.finite(cap_mg), " (capped at 1 g)", ""))
  ) |>
  knitr::kable(
    caption = paste(
      "Poster P-444 Table 2. Every regimen is a 3-hour intravenous infusion at",
      "a 1:1 sulbactam:durlobactam ratio. Birth is defined as 7 days",
      "post-natal."
    )
  )
```

The sources on disk give the regimens **by age**, but the model is driven by
**weight** --- and neither the CDC growth charts nor the Rhodin fat-free-mass
formula that the poster used to turn age into body size and renal function are
reported in them. Rather than invent an age-to-weight mapping, everything below
is presented against body weight, which is what the model actually sees. See
*Assumptions and deviations*.

## The 1 g cap produces a predictable inflection in exposure

At steady state, exposure over one dosing interval is exactly dose / clearance.
For an uncapped mg/kg dose that is
$\mathrm{WT}\,/\,\mathrm{WT}^{0.75} = \mathrm{WT}^{0.25}$ --- slowly rising with
size. Once the 1 g cap binds (at 1000 / 25 = 40 kg) the dose is constant and
exposure falls as $\mathrm{WT}^{-0.75}$. The model therefore predicts a peak in
exposure at exactly 40 kg, which is visible as the hump in the poster's own
Figure 2[A].

```{r cap-inflection}
wt_grid <- tibble::as_tibble(c(
  list(id = seq_len(41L), WT = seq(5, 65, length.out = 41L)),
  POSTER_COVS
)) |>
  dplyr::mutate(dose_mg = pmin(25 * WT, 1000))

# q6h out to 48 h so BOTH Day 1 (what the poster plots) and a steady-state
# dosing interval (where the allometry is exact) come from one solve.
cap_events <- dplyr::bind_rows(
  wt_grid |>
    tidyr::crossing(time = seq(0, 42, by = 6), cmt = c("central", "central_sbt")) |>
    dplyr::mutate(evid = 1L, amt = dose_mg, rate = dose_mg / 3,
                  dvid = NA_integer_),
  wt_grid |>
    tidyr::crossing(time = seq(0, 48, by = 0.05)) |>
    dplyr::mutate(evid = 0L, amt = NA_real_, rate = NA_real_,
                  cmt = "central", dvid = 1L)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

cap_sim <- solve_typical(mod_allo, cap_events, keep = c("WT", "dose_mg"))

cap_auc <- cap_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(id, WT, dose_mg) |>
  dplyr::summarise(
    Durlobactam = trapz(time[time <= 24], Cc[time <= 24]),
    Sulbactam   = trapz(time[time <= 24], Cc_sbt[time <= 24]),
    ss_dur      = trapz(time[time >= 42], Cc[time >= 42]),
    .groups     = "drop"
  )

peak_wt <- cap_auc$WT[which.max(cap_auc$Durlobactam)]

cap_auc |>
  tidyr::pivot_longer(c(Durlobactam, Sulbactam),
                      names_to = "Analyte", values_to = "auc") |>
  ggplot(aes(WT, auc, colour = Analyte)) +
  geom_vline(xintercept = 40, linetype = "dashed", colour = "grey40") +
  geom_line(linewidth = 0.9) +
  annotate("text", x = 40, y = Inf, label = "1 g cap binds (40 kg)",
           hjust = -0.05, vjust = 1.6, size = 3.2, colour = "grey30") +
  labs(
    x = "Body weight (kg)", y = expression(AUC[0-24]~(mg%.%h/L)),
    title = "Typical-value Day 1 exposure, 25 mg/kg q6h capped at 1 g",
    caption = paste(
      "Rising as WT^0.25 below the cap and falling as WT^-0.75 above it.",
      "Compare with the hump in poster P-444 Figure 2[A]."
    )
  )

below <- cap_auc |> dplyr::filter(WT <= 40)
above <- cap_auc |> dplyr::filter(WT >= 40)

slope <- function(df, col) {
  stats::coef(stats::lm(log(df[[col]]) ~ log(df$WT)))[[2]]
}

slopes <- tibble::tibble(
  "Weight range"          = rep(c("<= 40 kg (mg/kg dosing)",
                                  ">= 40 kg (1 g cap)"), 2),
  "Exposure measure"      = rep(c("Steady-state AUC over one interval",
                                  "Day 1 AUC0-24"), each = 2),
  "Fitted log-log slope"  = c(slope(below, "ss_dur"), slope(above, "ss_dur"),
                              slope(below, "Durlobactam"),
                              slope(above, "Durlobactam")),
  "Allometric prediction" = c(0.25, -0.75, NA, NA)
)

knitr::kable(
  slopes, digits = 4,
  caption = paste(
    "Durlobactam exposure versus body weight. At steady state AUC over one",
    "dosing interval is exactly dose / clearance, so the slopes are exactly",
    "the allometric predictions. Day 1 AUC0-24 is shallower below the cap and",
    "steeper above it because accumulation is not yet complete at 24 h and the",
    "elimination rate constant itself falls with weight (kel = CL / Vc scales",
    "as WT^-0.25), so a larger child retains proportionally more drug at 24 h."
  )
)
```

```{r cap-assert}
# EXACT gate on the quantity that is exact: at steady state AUCtau = dose / CL,
# so the two limbs must have the allometric exponents to within trapezoidal
# integration error.
stopifnot(
  abs(slope(below, "ss_dur") -  0.25) < 0.005,
  abs(slope(above, "ss_dur") - -0.75) < 0.005
)

# The peak sits on the cap weight. Grid spacing is 1.5 kg, so the nearest grid
# point to 40 is the tightest this can be.
stopifnot(abs(peak_wt - 40) <= 1.5)

# Day 1 is a derived, non-exact quantity; gate its SHAPE (rising then falling,
# with the deviations from 0.25 / -0.75 in the direction incomplete
# accumulation predicts) rather than pinning it to one run's value.
stopifnot(
  slope(below, "Durlobactam") > 0.15,
  slope(below, "Durlobactam") < 0.25,
  slope(above, "Durlobactam") < -0.75,
  slope(above, "Durlobactam") > -0.95
)
```

# Comparing the two approaches (cf. Figures 2 and 3)

The poster's headline modelling result is that the **Allometry + CL~CR~** model
predicts *lower* exposures than **Adult with Allometry**, which is why the
latter was chosen as the more conservative basis for dose selection.

```{r two-model-compare}
two_sim <- dplyr::bind_rows(
  solve_typical(mod_allo, cap_events, keep = c("WT", "dose_mg")) |>
    dplyr::mutate(model = "Adult with Allometry"),
  solve_typical(mod_crcl, cap_events, keep = c("WT", "dose_mg")) |>
    dplyr::mutate(model = "Allometry + CLcr")
)

two_auc <- two_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(model, id, WT) |>
  dplyr::summarise(
    Durlobactam = trapz(time[time <= 24], Cc[time <= 24]),      # Day 1, as plotted
    Sulbactam   = trapz(time[time <= 24], Cc_sbt[time <= 24]),
    ss_dur      = trapz(time[time >= 42], Cc[time >= 42]),      # steady-state interval
    ss_sbt      = trapz(time[time >= 42], Cc_sbt[time >= 42]),
    vc          = dplyr::first(vc),
    .groups     = "drop"
  )

two_auc |>
  tidyr::pivot_longer(c(Durlobactam, Sulbactam),
                      names_to = "Analyte", values_to = "auc") |>
  ggplot(aes(WT, auc, colour = model)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~Analyte, scales = "free_y") +
  labs(
    x = "Body weight (kg)", y = expression(AUC[0-24]~(mg%.%h/L)),
    colour = NULL,
    title = "Typical-value Day 1 exposure under the poster's two approaches",
    caption = "Compare with poster P-444 Figure 2 (upper) versus Figure 3 (lower)."
  ) +
  theme(legend.position = "bottom")
```

```{r two-model-ratio}
wide <- two_auc |>
  dplyr::select(model, id, WT, Durlobactam, Sulbactam, ss_dur, ss_sbt) |>
  tidyr::pivot_longer(c(Durlobactam, Sulbactam, ss_dur, ss_sbt),
                      names_to = "measure", values_to = "auc") |>
  tidyr::pivot_wider(names_from = model, values_from = auc) |>
  dplyr::mutate(ratio = `Allometry + CLcr` / `Adult with Allometry`)

ratio_tab <- wide |>
  dplyr::group_by(measure) |>
  dplyr::summarise(ratio = stats::median(ratio), .groups = "drop") |>
  dplyr::mutate(
    Analyte  = ifelse(grepl("dur|Durlo", measure), "Durlobactam", "Sulbactam"),
    Exposure = ifelse(grepl("^ss_", measure),
                      "Steady-state AUC over one interval", "Day 1 AUC0-24"),
    Exact    = ifelse(grepl("^ss_", measure),
                      ifelse(grepl("dur", measure), 1, 1 - 0.444), NA_real_)
  )

ratio_tab |>
  dplyr::select(Analyte, Exposure,
                "Median ratio (CLcr model / allometry model)" = ratio,
                "Exact algebraic value" = Exact) |>
  dplyr::arrange(Analyte, Exposure) |>
  knitr::kable(
    digits  = 4,
    caption = paste(
      "Poster P-444 Results: 'Compared to the adult with allometry model, the",
      "adult with allometry + CLCR model predicted lower SUD exposures.' At a",
      "common CLcr, the steady-state ratio is exact algebra; the Day 1 ratio",
      "is not, because the two models also differ in central volume."
    )
  )
```

At steady state, exposure over one dosing interval is dose / clearance and does
not depend on volume at all, which makes the two steady-state ratios exact:

* Durlobactam's bacteremia effect acts on central *volume* (+3.32), not
  clearance, and durlobactam carries no region or infection effect on
  clearance. Removing those covariates leaves steady-state durlobactam exposure
  **exactly unchanged** --- ratio 1.
* Sulbactam's bacteremia effect acts on *clearance* (-0.444). Removing it raises
  clearance by 1 / (1 - 0.444), so exposure falls by exactly 0.556.

```{r two-model-assert}
ss <- ratio_tab |> dplyr::filter(grepl("^ss_", measure))
stopifnot(nrow(ss) == 2L, max(abs(ss$ratio - ss$Exact)) < 1e-3)

# The volume ratio is likewise exact: 1 + 3.32 = 4.32.
vc_ratio <- two_auc |>
  dplyr::select(model, id, vc) |>
  tidyr::pivot_wider(names_from = model, values_from = vc) |>
  dplyr::summarise(r = stats::median(`Adult with Allometry` / `Allometry + CLcr`)) |>
  dplyr::pull(r)
stopifnot(abs(vc_ratio - 4.32) < 1e-6)

# The poster's claim -- lower exposures under Allometry + CLcr -- is reproduced
# for sulbactam on BOTH measures.
stopifnot(all(ratio_tab$ratio[ratio_tab$Analyte == "Sulbactam"] < 0.7))
```

**Durlobactam does not reproduce the poster's direction, and that is
informative rather than a defect.** On Day 1 the ratio comes out *above* 1: the
"Adult with Allometry" model gives durlobactam a 4.32-fold larger central
volume, so its elimination rate constant `kel = CL / Vc` is 4.32-fold smaller
and proportionally more drug is still in the body at 24 h, depressing its
AUC~0-24~. At steady state that transient washes out and the ratio returns to
exactly 1. Either way, at a **common** CL~CR~ the two models cannot make
durlobactam exposure lower under "Allometry + CL~CR~", because durlobactam
clearance is identical between them. Since the poster's Figure 3[A] does show
lower durlobactam exposure, the two models must have been driven with
**different renal function** --- which is what their names imply and what the
Rhodin formula in the abstract was for. That formula is not in either source on
disk, so this vignette does not guess at it. See *Assumptions and deviations*.

# Virtual cohort and stochastic simulation

```{r cohort}
set.seed(20260910)

N_PER_ARM <- 150L   # under the 200-per-arm validation-vignette cap

# Weight bands spanning the pediatric range. These are illustrative body-weight
# strata, NOT an age-to-weight mapping taken from the sources -- see
# *Assumptions and deviations*.
bands <- tibble::tribble(
  ~band,        ~wt_lo, ~wt_hi, ~mg_per_kg, ~tau,
  "2-5 kg",       2,      5,     20,         12,
  "5-10 kg",      5,     10,     25,          6,
  "10-20 kg",    10,     20,     25,          6,
  "20-40 kg",    20,     40,     25,          6,
  "40-65 kg",    40,     65,     25,          6
)

cohort <- bands |>
  dplyr::mutate(row = dplyr::row_number()) |>
  tidyr::crossing(k = seq_len(N_PER_ARM)) |>
  dplyr::mutate(
    id      = (row - 1L) * 1000L + k,
    WT      = stats::runif(dplyr::n(), wt_lo, wt_hi),
    dose_mg = pmin(mg_per_kg * WT, 1000)
  ) |>
  dplyr::select(id, band, tau, WT, dose_mg)

for (nm in names(POSTER_COVS)) cohort[[nm]] <- POSTER_COVS[[nm]]

cohort_events <- dplyr::bind_rows(
  # Dosing times differ by band (q6h or q12h), so cross with every hour and
  # keep the multiples of that subject's interval.
  cohort |>
    tidyr::crossing(time = 0:23) |>
    dplyr::filter(time %% tau == 0) |>
    tidyr::crossing(cmt = c("central", "central_sbt")) |>
    dplyr::mutate(evid = 1L, amt = dose_mg, rate = dose_mg / 3,
                  dvid = NA_integer_),
  cohort |>
    tidyr::crossing(time = seq(0, 24, by = 0.25)) |>
    dplyr::mutate(evid = 0L, amt = NA_real_, rate = NA_real_,
                  cmt = "central", dvid = 1L)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(nrow(cohort) == 5L * N_PER_ARM)
```

```{r simulate}
sim <- rxode2::rxSolve(
  mod_allo, events = cohort_events,
  keep = c("band", "WT", "dose_mg"),
  useLinCmt = FALSE
) |>
  as.data.frame()

dplyr::glimpse(sim[, c("id", "time", "Cc", "Cc_sbt", "Celf", "Celf_sbt",
                       "cl", "cl_sbt", "band", "WT")])
```

## Simulated pediatric exposures against the adult reference range

Poster P-444 Table 1 gives the adult median and 5th-95th percentile exposures
that the pediatric regimens were targeted to sit inside. The poster's stated
result is that

> The adult with allometry model provided SUD exposures that were predominantly
> contained within the 5th and 95th percentiles of adult exposures based on
> AUC~0-24~

```{r adult-table1}
adult_exposure <- tibble::tribble(
  ~Analyte,      ~Parameter,   ~day, ~median, ~p05,  ~p95,
  "Sulbactam",   "AUC0-24",     1,    441,     196,   1170,
  "Sulbactam",   "AUC0-24",     3,    469,     192,   1710,
  "Sulbactam",   "Cmax",        1,     30.0,    14.7,   75.9,
  "Sulbactam",   "Cmax",        3,     29.5,    14.3,   88.8,
  "Durlobactam", "AUC0-24",     1,    442,     277,    904,
  "Durlobactam", "AUC0-24",     3,    479,     239,   1180,
  "Durlobactam", "Cmax",        1,     30.3,    18.4,   58.9,
  "Durlobactam", "Cmax",        3,     30.5,    17.3,   65.0
)

adult_exposure |>
  dplyr::transmute(
    Analyte, Parameter, Day = day,
    "Median (5th-95th percentile)" =
      paste0(median, " (", p05, "-", p95, ")")
  ) |>
  knitr::kable(
    caption = paste(
      "Poster P-444 Table 1: adult sulbactam and durlobactam exposures on Day 1",
      "and Day 3. AUC0-24 in h*mg/L, Cmax in mg/L."
    )
  )
```

```{r envelope-check}
env <- adult_exposure |>
  dplyr::filter(Parameter == "AUC0-24", day == 1) |>
  dplyr::select(Analyte, p05, p95)

ped_auc <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(band, id) |>
  dplyr::summarise(Durlobactam = trapz(time, Cc),
                   Sulbactam   = trapz(time, Cc_sbt), .groups = "drop") |>
  tidyr::pivot_longer(c(Durlobactam, Sulbactam),
                      names_to = "Analyte", values_to = "auc") |>
  dplyr::left_join(env, by = "Analyte")

band_summary <- ped_auc |>
  dplyr::group_by(Analyte, band, p05, p95) |>
  dplyr::summarise(
    med = stats::median(auc),
    pct_inside = 100 * mean(auc >= p05 & auc <= p95),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    median_inside = med >= p05 & med <= p95,
    # The 2-5 kg band is the neonatal stratum, where BSA-normalized renal
    # function is far below the adult reference. Reproducing it needs the
    # Rhodin maturation function, which is not in either source on disk, so
    # it is recorded as a deviation rather than gated. See Errata.
    deviation = band == "2-5 kg"
  )

band_summary |>
  dplyr::transmute(
    Analyte,
    "Weight band"             = band,
    "Median AUC0-24"          = med,
    "% inside adult 5th-95th" = pct_inside,
    "Median inside adult range" = ifelse(median_inside, "yes", "no"),
    "Deviation"               = ifelse(deviation, "yes", "")
  ) |>
  knitr::kable(
    digits  = 1,
    caption = paste(
      "Simulated Day 1 AUC0-24 (h*mg/L) by weight band under the 'Adult with",
      "Allometry' model, against the adult 5th-95th percentile range of poster",
      "P-444 Table 1. The neonatal 2-5 kg band is flagged as a deviation: it",
      "needs the renal-maturation function this vignette cannot reproduce."
    )
  )
```

```{r envelope-assert}
gated <- band_summary |> dplyr::filter(!deviation)

# Guard against a vacuous gate: confirm there are rows to test (all(logical(0))
# is TRUE, so a label mismatch would silently pass).
stopifnot(nrow(gated) == 8L)

# The poster's claim is about the CENTRE of the pediatric distribution sitting
# inside the adult range, so assert on band medians -- a cohort's extremes are
# not reproducible across rxode2 versions or solver thread counts.
stopifnot(all(gated$median_inside))

# "Predominantly contained" is qualitative; require a clear majority per band
# rather than a threshold read off one run.
stopifnot(all(gated$pct_inside > 55))
```

The four bands from 5 kg upward reproduce the poster's headline result: the
median simulated pediatric exposure sits inside the adult 5th-95th percentile
range for both analytes, and the large majority of individual subjects do too.

The neonatal **2-5 kg** band falls below the adult range for both analytes,
and this is expected rather than a transcription error. That band is dosed
20 mg/kg q12h (two doses a day rather than four), and the model as packaged
scales clearance by weight alone --- so at a fixed CL~CR~ of 100
mL/min/1.73 m^2 it gives a term neonate a *fully mature* renal clearance per
kilogram. The poster instead assigned each simulated subject a Rhodin-derived
renal function, which in a neonate is a fraction of the adult value. The
sensitivity below shows how far renal function has to fall to close the gap.

```{r neonatal-sensitivity}
neo <- tibble::as_tibble(c(
  list(id = seq_len(6L), WT = 3.5),
  utils::modifyList(POSTER_COVS, list(CRCL = NULL, RENALIMP_SEV = NULL))
)) |>
  dplyr::mutate(
    CRCL         = c(100, 75, 50, 30, 20, 10),
    RENALIMP_SEV = as.numeric(CRCL < 30)
  )

neo_ev  <- make_events(neo, dose_mg = 20 * 3.5, tau = 12, tmax = 24, by = 0.05)
neo_sim <- solve_typical(mod_allo, neo_ev, keep = c("WT", "CRCL"))

neo_tab <- neo_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(CRCL) |>
  dplyr::summarise(Durlobactam = trapz(time, Cc),
                   Sulbactam   = trapz(time, Cc_sbt), .groups = "drop") |>
  # dplyr returns groups in ascending CRCL order; make that explicit so the
  # monotonicity assertion below cannot silently invert.
  dplyr::arrange(CRCL) |>
  dplyr::mutate(
    `Durlobactam inside adult range` =
      ifelse(Durlobactam >= 277 & Durlobactam <= 904, "yes", "no"),
    `Sulbactam inside adult range` =
      ifelse(Sulbactam >= 196 & Sulbactam <= 1170, "yes", "no")
  )

neo_tab |>
  dplyr::rename("CLcr (mL/min/1.73 m^2)" = CRCL) |>
  knitr::kable(
    digits  = 1,
    caption = paste(
      "Typical-value Day 1 AUC0-24 (h*mg/L) for a 3.5 kg term neonate on",
      "20 mg/kg q12h, as a function of the renal function assigned. The adult",
      "reference ranges are 277-904 (durlobactam) and 196-1170 (sulbactam).",
      "Exposure enters the adult range once renal function drops well below",
      "the adult reference, which is what a maturation function supplies."
    )
  )

# Deterministic typical-value monotonicity. The table is sorted by ASCENDING
# CLcr, so exposure must be strictly DECREASING down the rows.
stopifnot(
  nrow(neo_tab) == 6L,
  all(diff(neo_tab$Durlobactam) < 0),
  all(diff(neo_tab$Sulbactam)   < 0),
  # At the adult reference CLcr the neonate falls below the adult range ...
  neo_tab$`Durlobactam inside adult range`[neo_tab$CRCL == 100] == "no",
  # ... and the gap IS closable by renal function alone, which is the point.
  any(neo_tab$`Durlobactam inside adult range` == "yes")
)
```

```{r envelope-figure}
ped_auc |>
  ggplot(aes(band, auc)) +
  geom_hline(aes(yintercept = p05), linetype = "dashed", colour = "grey40") +
  geom_hline(aes(yintercept = p95), linetype = "dashed", colour = "grey40") +
  geom_boxplot(outlier.size = 0.5, fill = "steelblue", alpha = 0.35) +
  facet_wrap(~Analyte, scales = "free_y") +
  labs(
    x = "Body-weight band", y = expression(AUC[0-24]~(mg%.%h/L)),
    title = "Simulated pediatric Day 1 exposure versus the adult reference range",
    caption = paste(
      "Dashed lines are the adult 5th and 95th percentiles of poster P-444",
      "Table 1. Compare with poster P-444 Figure 2[A] and [B]."
    )
  ) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

# PKNCA validation

```{r pknca}
#' Run PKNCA over 0-24 h for one concentration column of a simulation frame.
nca_for <- function(sim_df, conc_col, dose_df) {
  sim_nca <- sim_df |>
    dplyr::rename(Cc_use = dplyr::all_of(conc_col)) |>
    dplyr::filter(!is.na(Cc_use)) |>
    dplyr::select(id, time, Cc = Cc_use, band)

  # Guarantee a time-zero row per (band, id); pre-dose IV concentration is 0.
  sim_nca <- dplyr::bind_rows(
    sim_nca,
    sim_nca |> dplyr::distinct(id, band) |> dplyr::mutate(time = 0, Cc = 0)
  ) |>
    dplyr::distinct(band, id, time, .keep_all = TRUE) |>
    dplyr::arrange(id, time)

  conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | band + id,
                               concu = "mg/L", timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | band + id, doseu = "mg")

  intervals <- data.frame(start = 0, end = 24,
                          cmax = TRUE, tmax = TRUE, auclast = TRUE)

  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}

dose_df <- cohort_events |>
  dplyr::filter(evid == 1, cmt == "central") |>
  dplyr::select(id, time, amt, band)

nca_dur <- nca_for(sim, "Cc", dose_df)
nca_sbt <- nca_for(sim, "Cc_sbt", dose_df)

stopifnot(nrow(as.data.frame(nca_dur)) > 0,
          nrow(as.data.frame(nca_sbt)) > 0)
```

```{r pknca-vs-trapz}
# Cross-check PKNCA against the independent trapezoidal AUC used above. Two
# implementations of the same integral over the same grid must agree closely.
pknca_auc <- as.data.frame(nca_dur) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(band, id, pknca = PPORRES)

chk <- ped_auc |>
  dplyr::filter(Analyte == "Durlobactam") |>
  dplyr::select(band, id, trapz = auc) |>
  dplyr::inner_join(pknca_auc, by = c("band", "id")) |>
  dplyr::mutate(pct_diff = 100 * (pknca - trapz) / trapz)

stopifnot(
  nrow(chk) == 5L * N_PER_ARM,
  max(abs(chk$pct_diff)) < 0.5
)

tibble::tibble(
  "Subjects compared"          = nrow(chk),
  "Max |% difference|"         = max(abs(chk$pct_diff)),
  "Median |% difference|"      = stats::median(abs(chk$pct_diff))
) |>
  knitr::kable(
    digits  = 4,
    caption = paste(
      "PKNCA auclast over 0-24 h against the independent trapezoidal integral,",
      "per simulated subject."
    )
  )
```

## Adult exposures on Day 1 versus Day 3 (cf. Table 1)

At 75 kg the pediatric model reduces to the adult model, so the packaged model
can reproduce the poster's own adult reference column. The poster's stated
conclusion from Table 1 is a **lack of accumulation** --- and a Day 3 / Day 1
ratio is insensitive to the (unreported) covariate mix of the adult post-hoc
population, which the absolute values are not.

Table 1 describes the **real adult Phase 2/3 population**, whose infection-type
mix the poster does not report --- so it is *not* the bacteremia setting used
for the pediatric simulations. Both settings are shown below, because the
contrast between them is itself a structural check.

```{r accumulation}
adult_setting <- function(bacteremia) {
  covs <- utils::modifyList(POSTER_COVS, list(DIS_BACTEREMIA = bacteremia))
  subj <- tibble::as_tibble(c(list(id = 1L, WT = 75), covs))
  s <- solve_typical(mod_allo, make_events(subj, 1000, 6, 72, by = 0.05),
                     keep = "WT") |>
    dplyr::filter(!is.na(Cc))
  d1 <- s$time <= 24
  d3 <- s$time >= 48
  tibble::tibble(
    setting  = if (bacteremia == 1) "Bacteremia (pediatric-simulation setting)"
               else "Uninfected reference",
    vc_dur   = s$vc[1],
    dur_d1   = trapz(s$time[d1], s$Cc[d1]),
    dur_d3   = trapz(s$time[d3], s$Cc[d3]),
    sbt_d1   = trapz(s$time[d1], s$Cc_sbt[d1]),
    sbt_d3   = trapz(s$time[d3], s$Cc_sbt[d3]),
    dur_cmax_d1 = max(s$Cc[d1]),      dur_cmax_d3 = max(s$Cc[d3]),
    sbt_cmax_d1 = max(s$Cc_sbt[d1]),  sbt_cmax_d3 = max(s$Cc_sbt[d3])
  )
}

acc <- dplyr::bind_rows(adult_setting(0), adult_setting(1))

acc_tab <- acc |>
  dplyr::transmute(
    Setting = setting,
    "Durlobactam Vc (L)"    = vc_dur,
    "Durlobactam AUC D3/D1" = dur_d3 / dur_d1,
    "Durlobactam Cmax D3/D1"= dur_cmax_d3 / dur_cmax_d1,
    "Sulbactam AUC D3/D1"   = sbt_d3 / sbt_d1,
    "Sulbactam Cmax D3/D1"  = sbt_cmax_d3 / sbt_cmax_d1
  )

dplyr::bind_rows(
  acc_tab,
  tibble::tibble(
    Setting = "Poster P-444 Table 1 (observed adults)",
    `Durlobactam Vc (L)`     = NA_real_,
    `Durlobactam AUC D3/D1`  = 479 / 442,
    `Durlobactam Cmax D3/D1` = 30.5 / 30.3,
    `Sulbactam AUC D3/D1`    = 469 / 441,
    `Sulbactam Cmax D3/D1`   = 29.5 / 30.0
  )
) |>
  knitr::kable(
    digits  = 3,
    caption = paste(
      "Day 3 / Day 1 accumulation ratios at the adult 1 g / 1 g q6h regimen.",
      "Poster P-444 Results: the Table 1 exposures 'demonstrate lack of",
      "accumulation'. The uninfected reference reproduces that; the bacteremia",
      "setting accumulates more because its 4.32-fold larger central volume",
      "lengthens the effective half-life."
    )
  )
```

```{r accumulation-assert}
ref  <- acc |> dplyr::filter(setting == "Uninfected reference")
bact <- acc |> dplyr::filter(setting != "Uninfected reference")
stopifnot(nrow(ref) == 1L, nrow(bact) == 1L)

# 1. At the uninfected reference the model shows the near-absence of
#    accumulation that Table 1 reports (published ratios 1.06-1.08 for AUC and
#    0.98-1.01 for Cmax).
stopifnot(
  abs(ref$dur_d3 / ref$dur_d1 - 1) < 0.10,
  abs(ref$sbt_d3 / ref$sbt_d1 - 1) < 0.10,
  abs(ref$dur_cmax_d3 / ref$dur_cmax_d1 - 1) < 0.02,
  abs(ref$sbt_cmax_d3 / ref$sbt_cmax_d1 - 1) < 0.02
)

# 2. Structural consequence of the bacteremia Vc shift: a larger volume at the
#    same clearance lengthens the effective half-life, so accumulation MUST be
#    greater. Deterministic typical values, so a plain inequality is safe.
stopifnot(
  bact$vc_dur / ref$vc_dur > 4,
  bact$dur_d3 / bact$dur_d1 > ref$dur_d3 / ref$dur_d1,
  bact$sbt_d3 / bact$sbt_d1 > ref$sbt_d3 / ref$sbt_d1
)

# 3. Steady state is a covariate-independent anchor: Day 3 exposure is
#    dose / CL regardless of volume, so the two settings must agree on
#    durlobactam Day 3 (durlobactam has no infection effect on clearance).
stopifnot(abs(bact$dur_d3 / ref$dur_d3 - 1) < 0.01)
```

```{r absolute-adult}
tibble::tibble(
  Analyte   = c("Durlobactam", "Durlobactam", "Sulbactam", "Sulbactam"),
  Parameter = c("AUC0-24 (Day 3)", "Cmax (Day 3)",
                "AUC0-24 (Day 3)", "Cmax (Day 3)"),
  Simulated = c(ref$dur_d3, ref$dur_cmax_d3, ref$sbt_d3, ref$sbt_cmax_d3),
  Published = c(479, 30.5, 469, 29.5)
) |>
  dplyr::mutate(`% difference` = 100 * (Simulated - Published) / Published) |>
  dplyr::rename("Poster Table 1 median" = Published) |>
  knitr::kable(
    digits  = 1,
    caption = paste(
      "Typical-value adult Day 3 exposures at 1 g / 1 g q6h (75 kg,",
      "non-East-Asian, CLcr 100, uninfected reference) against the poster's",
      "Table 1 medians. This is a DEVIATION table, not a gate. Durlobactam is",
      "close; sulbactam is under-predicted because every infection-type",
      "covariate in the adult model REDUCES sulbactam clearance by 16-44%, and",
      "the Table 1 population is entirely infected patients whose mix the",
      "poster does not report."
    )
  )
```

# Assumptions and deviations

- **The poster, not the abstract, is the model source.** The PMC-indexed record
  (`PMC12792777`) is a one-page IDWeek abstract containing no parameter values,
  no equations, and no covariate structure --- its "Table 1" is referenced but
  not reproduced in the supplement. The publicly posted sponsor poster P-444
  supplies the complete specification. Both are cited in the model files.

- **Author order differs between abstract and poster.** Abstract: Larson,
  O'Donnell, Tanudra, Cammarata, Rubino. Poster: Cammarata, Larson, Tanudra,
  O'Donnell, Bhavnani, Rubino (the poster also adds Bhavnani). The files are
  named `Larson_2026_*` after the indexed abstract. Note the parent adult model
  in this library is `Cammarata_2024_sulbactam_durlobactam` --- the two are
  different publications by an overlapping author group, not duplicates.

- **Reference weight 75 kg is inherited, not restated.** The poster says only
  that the exponents changed; it never writes the covariate equation or names a
  reference weight. 75 kg is the adult model's reference subject weight
  (Cammarata 2024 Results: "have a body weight of 75 kg"), and the poster
  frames this model as the adult model "with the only change being the use of
  allometric scaling with fixed exponents". A different centring constant
  (70 kg is the more common allometric convention) would rescale every typical
  value by `(75/70)^k`, i.e. by 5.4% on clearance and 7.1% on volume.

- **Q and Vp exponents are assigned by kind, not stated numerically.** The
  poster gives 0.75 "for clearance" and 1.0 "for volume of distribution", and
  separately notes that "WTKG effects added to Q and Vp for both drugs". Q is a
  distributional *clearance* and Vp a *volume*, so they take 0.75 and 1.0
  respectively. The poster does not print these two numbers against those two
  parameter names.

- **The Rhodin fat-free-mass renal-maturation formula is not reproduced.** The
  abstract states that "Renal function was estimated using the Rhodin formula
  using free fat mass", but neither source on disk gives that formula or its
  constants, and the poster describes the cohort simply as "assuming normal
  renal function". Every simulation here therefore sets
  `CRCL = 100 mL/min/1.73 m^2` --- the model's own definition of normal renal
  function (Cammarata 2024 Results) --- rather than importing constants from an
  off-disk publication. **Consequence:** the CLcr term is at its reference value
  and contributes a factor of exactly 1, so the two packaged models differ here
  only through the covariates that model 2 removes. In the poster's own figures
  the two models differ for durlobactam as well, which they cannot do at a
  common CLcr because durlobactam AUC depends only on clearance and durlobactam
  carries no infection-type or region effect on clearance. The most likely
  explanation is that the poster fed per-subject Rhodin-derived CLcr to the
  "Allometry + CL~CR~" model and held renal function at the adult reference for
  "Adult with Allometry" --- which is what the two model names suggest --- but
  the poster does not say so, and this vignette does not guess.

- **Documented deviation: the neonatal weight band.** With renal function held
  at the adult reference, simulated exposure in the 2-5 kg band falls below the
  adult 5th percentile for both analytes, whereas the poster's Figure 2 shows
  its Cohort 5 medians inside the adult range. The mechanism is the missing
  renal maturation, not a transcription error: the packaged model scales
  clearance by weight alone, so at `CRCL = 100` a term neonate is given a fully
  mature renal clearance per kilogram. The sensitivity table above shows
  exposure re-entering the adult range once renal function is reduced to
  neonatal levels. This band is reported in the results table but excluded from
  the gate rather than having the gate widened until it passed.

- **Age-to-weight mapping is not reproduced.** Poster Table 2 gives the
  regimens by age cohort, and the cohort was built from CDC growth charts,
  which are not on disk. Rather than invent a weight-for-age table, results are
  presented against body weight, which is the model's actual covariate. The
  weight bands in the stochastic cohort are illustrative strata chosen to span
  the pediatric range; they are **not** taken from any source and no claim is
  made that a given band corresponds to a given Table 2 age cohort. The one
  age-related quantity that *is* reproduced is structural: the 1 g cap binds at
  40 kg, and the resulting exposure peak is visible in the poster's Figure 2[A].

- **Simulation covariate settings come from the poster's Figure 1 notes:**
  infection type set to bacteremia, East Asian flag set to not East Asian,
  normal renal function. Bacteremia carries the largest covariate effects in the
  adult model (durlobactam Vc +3.32, sulbactam CL -0.444), so this choice raises
  simulated exposure and is part of why "Adult with Allometry" is the more
  conservative approach. The study-phase indicators are set to Phase 3, which
  selects the largest durlobactam proportional residual magnitude; residual
  error affects only `sim`-level noise, not `Cc`, and none of the gates above
  use residual-error-carrying quantities.

- **PK/PD target attainment is not reproduced.** The poster reports PTA against
  50% fT>MIC (sulbactam) and a free-drug AUC/MIC ratio of 10 (durlobactam), but
  gives neither the MIC distribution nor the protein-binding values used for
  the free-drug conversion in the pediatric analysis. PTA curves (Figure 2[C],
  [D] and Figure 3[C], [D]) are therefore out of scope here. The ELF
  observables the poster used for the durlobactam target *are* packaged and
  available as `Celf` / `Celf_sbt`.

- **Absolute adult exposures are a deviation table, not a gate.** Poster Table 1
  reports medians and 5th-95th percentiles of post-hoc exposures over the real
  adult Phase 2/3 population, whose covariate mix (weight, CLcr, infection-type
  proportions) is not reported. A single typical-value solve cannot be expected
  to land on that median, so the accumulation *ratio* --- which is insensitive
  to the covariate mix --- is what is asserted.

- **Inherited-value caveats carry over from the parent model.** In particular,
  the fraction excreted renally (FE) is the Table 1 value of 0.479, which
  Cammarata 2024 notes was subsequently revised to 0.660 / 0.648 in a refit
  whose parameter table is reported only as "data not shown"; and the published
  adult post-hoc exposures were generated with that unpublished refitted model.
  See the `Cammarata_2024_sulbactam_durlobactam` vignette for the full
  discussion.

- **Errata.** No erratum, corrigendum or author correction to
  doi:10.1093/ofid/ofaf695.659 was found. One internal inconsistency is worth
  recording: the poster labels its two Cohort 5 sub-rows "Cohort 5, Subgroup 1"
  twice, where the second is evidently Subgroup 2 (the birth-to-<2-months
  stratum). The regimen table above uses the age ranges, which are unambiguous.

- **Non-paper-derived values.** None. Every `ini()` value is either printed on
  poster P-444 (the eight allometric exponents) or inherited unchanged from the
  already-source-traced `Cammarata_2024_sulbactam_durlobactam` model, and the
  inheritance is verified mechanically in the *Source trace* section above. No
  value was digitised from a figure, obtained by correspondence, or taken from
  an off-disk publication.
