---
title: "TBAJ-587 (Leding 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TBAJ-587 (Leding 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Leding_2026_tbaj587"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1002/bcp.70333>
* Supplement (Code S1 final NONMEM control stream, Tables S1-S2, Figures S1-S5):
  <https://doi.org/10.1002/bcp.70333>, Supporting Information

TBAJ-587 is a second-in-class diarylquinoline (DARQ) antitubercular, the class
founded by bedaquiline. Leding 2026 describes the single-ascending-dose
pharmacokinetics of TBAJ-587 and its two main metabolites, M2 and M3, and uses
the model to pick a phase 2a dose and to ask whether a loading-dose period is
worth its complexity.

The extraction is **unusually well constrained** because the supplement carries
the complete final NONMEM control stream (Code S1). Two facts that only the
control stream settles, and that would each have been a silent error if guessed
from the paper body:

1. **`ktr = (NN + 1) / MTT`**, not `NN / MTT`. Code S1 states
   `KTR = (NN+1)/MTT` outright. The paper body gives only "absorption transit
   compartments" and an estimated `NN`, and the model estimates a *separate*
   `ka`, which is the configuration in which the house convention of some
   groups is `n / MTT`. Guessing wrong shifts `ktr` by 42% here.
2. **The metabolite formation flux is not subtracted from the parent.** Code S1
   keeps `-K20*A(2) = -CL/V2*A(2)` in `DADT(2)` in full, and adds
   `K25 = CL/V2*FM3` and `K28 = CL/V2*FM2` as inflows to the two metabolite
   central compartments with no compensating loss. See "Parallel formation"
   below, which is the single most important thing to understand about this
   model.

One equally important trap runs the other way: **the Code S1 `$THETA` and
`$OMEGA` records for the metabolite sub-models are *initial* estimates, not
final ones.** `CLM3` is 20.121 in Code S1 against 18.5 in Table 1, and `Q3`
is 17.4191 against 9.09 -- a factor of 1.9. The parent block, by contrast, is
`FIX` at exactly its Table 1 values. So this extraction takes **structure and
every equation form from Code S1, and every final value from Table 1**.

# Population

```{r population, echo = FALSE}
pop <- ui$population
tibble::tibble(Field = names(pop),
               Value = vapply(pop, function(x) paste(x, collapse = "; "), character(1))) |>
  knitr::kable(caption = "Population metadata (Leding 2026 Methods 2.1, Results 3.1).")
```

Forty-two healthy volunteers (male or female of non-childbearing potential,
18-64 years, BMI 15.5-32.0 kg/m^2, minimum weight 50.0 kg) received a single
oral dose of TBAJ-587 in trial NCT04890535 Part 1 at one Netherlands site.
Six single-ascending-dose cohorts received 25, 50, 100, 200, 400 and 800 mg
(n = 6, 4, 5, 8, 4 and 6), and a separate nine-subject food-effect cohort
received 200 mg with a high-calorie high-fat meal. Plasma was sampled densely
to 312 h, and 29 of 42 subjects (69%) were followed to Day 98 with 23 of those
29 (79%) reaching Day 126 -- follow-up that matters here because the terminal
half-life runs to months. 1929 observations were available for each of the three
analytes. The sex split is not reported in the paper, so `sex_female_pct` is
deliberately absent from the metadata rather than guessed.

Note that the food effect is **between-subject**, not a crossover: the fed data
come from nine subjects dosed only at 200 mg. Leding 2026 Discussion is explicit
that the fed simulations therefore assume the dose dependence of apparent oral
clearance, which was driven by the six fasted dose levels, carries over to the
fed state.

# Source trace

Per-parameter provenance is recorded as an in-file comment beside each `ini()`
entry in `inst/modeldb/specificDrugs/Leding_2026_tbaj587.R`. Collected here for
review.

| Equation / parameter | Value | Source location |
|---|---|---|
| Absorption input `trin` (Savic gamma density, Stirling `LNFAC`) | n/a | Code S1 `$DES` `DADT(1)`; `LNFAC` defined in `$PK` |
| `ktr = (nn + 1) / mtt` | n/a | Code S1 `$PK` `KTR = (NN+1)/MTT` |
| `f(depot) <- 0` | n/a | Code S1 `$PK` `F1 = 0` |
| Parent 3-cmt disposition | n/a | Code S1 `$DES` `DADT(2)`-`DADT(4)`; Figure 2; Results 3.2 |
| M3 3-cmt disposition | n/a | Code S1 `$DES` `DADT(5)`-`DADT(7)`; Figure 2; Results 3.2 |
| M2 2-cmt disposition | n/a | Code S1 `$DES` `DADT(8)`-`DADT(9)`; Figure 2; Results 3.2 |
| Parallel formation `K25`, `K28` (no parent loss term) | n/a | Code S1 `$PK` `K25 = CL/V2*FM3`, `K28 = CL/V2*FM2`; Methods 2.2; Table 1 footnote a |
| `lfdepot` (F fasted) | 1 (fixed) | Table 1 `F fasted` = 1 FIX; Code S1 `$THETA(2)` |
| `lka` | 0.0866 1/h | Table 1 `ka`, RSE 3.5% |
| `lcl` | 5.40 L/h | Table 1 `CL TBAJ-587`, RSE 8.4% |
| `lvc` | 88.2 L | Table 1 `V C,TBAJ-587`, RSE 6.8% |
| `lq` / `lvp` | 31.1 L/h / 1910 L | Table 1 `Q P1,TBAJ-587`, `V P1,TBAJ-587` |
| `lq2` / `lvp2` | 31.5 L/h / 22 500 L | Table 1 `Q P2,TBAJ-587`, `V P2,TBAJ-587` |
| `lmtt` | 0.837 h | Table 1 `MTT fasted`, RSE 4.5% |
| `lnn` | 2.36 | Table 1 `NN`, RSE 8% |
| `e_fed_highfat_fdepot` | 0.688 | Table 1 `Fed state on F fasted`; form from footnote d |
| `e_dose_cl` | 0.298 | Table 1 `Dose covariate on CL TBAJ-587`; power form from footnote e |
| `e_dose_ka` | -0.000462 1/mg | Code S1 `$THETA(13)`; Table 1 prints -0.00046 rounded; exponential form from footnote f |
| `e_fed_highfat_mtt` | 0.958 | Table 1 `Fed state on MTT fasted`; form from footnote d |
| `lfm_m3` | 1 (fixed) | Table 1 `f m,M3` = 1 FIX; Code S1 `$THETA(15)` |
| `lcl_m3` / `lvc_m3` | 18.5 L/h / 816 L | Table 1 `CL M3`, `V C,M3` |
| `lq_m3` / `lvp_m3` | 9.09 L/h / 1630 L | Table 1 `Q P1,M3`, `V P1,M3` |
| `lq2_m3` / `lvp2_m3` | 99.6 L/h / 3010 L | Table 1 `Q P2,M3`, `V P2,M3` |
| `e_dose_fm_m3` | -0.418 | Table 1 `Dose covariate on f m,M3`; power form from footnote e |
| `e_fed_highfat_fm_m3` | -0.479 | Table 1 `Fed state on f m,M3`; form from footnote d |
| `lfm_m2` | 1 (fixed) | Table 1 `f m,M2` = 1 FIX; Code S1 `$THETA(24)` |
| `lcl_m2` / `lvc_m2` | 34.1 L/h / 247 L | Table 1 `CL M2`, `V C,M2` |
| `lq_m2` / `lvp_m2` | 174 L/h / 12 200 L | Table 1 `Q P1,M2`, `V P1,M2` |
| `e_dose_fm_m2` | -0.373 | Table 1 `Dose covariate on f m,M2`; power form from footnote e |
| `e_fed_highfat_fm_m2` | -0.547 | Table 1 `Fed state on f m,M2`; form from footnote d |
| `e_dose_cl_m2` | 0.146 | Table 1 `Dose covariate on CL M2`; power form from footnote e |
| Parent IIV (7 etas) | see model file | Code S1 `$OMEGA(1)`-`$OMEGA(7)`, all FIX; each reproduces its Table 1 CV% |
| Metabolite IIV (5 etas) | see model file | Table 1 IIV CV% column, back-transformed as `omega^2 = log(1 + CV^2)` (footnote c) |
| `expSd` / `expSd_m3` / `expSd_m2` | 0.1836 / 0.1954 / 0.1634 | `sqrt()` of the Table 1 variances 0.0337 / 0.0382 / 0.0267; `lnorm` form from Code S1 `$ERROR` |

## Parallel formation with both fractions anchored at 1

This is the structural feature most likely to be mis-transcribed, so it is worth
stating plainly. Methods 2.2 develops the metabolites "assuming parallel
formation, using the fixed TBAJ-587 PK model and the assumption of f m, the
fraction of TBAJ-587 cleared to metabolite, equalling one", and Table 1 footnote
a adds that for the metabolites "clearances and volumes are additionally
relative to unknown fractions metabolized".

In Code S1 that becomes: the parent's total elimination is exactly `CL/V2`, and
**each** metabolite separately receives an inflow of `fm * CL/V2 * A(2)` with
`fm` fixed at 1. The two `fm` values therefore anchor at 1 rather than summing
to 1, the metabolite amounts are not mass-conserved against the parent, and each
metabolite's true (unidentifiable) fraction metabolised is absorbed into its own
apparent clearance and volumes. The only estimated content of the two `fm`
parameters is their dose and food covariate effects, plus IIV on `fm_m2`.

The tempting "repair" -- subtracting the formation fluxes from the parent to
balance mass -- would silently rescale every apparent metabolite parameter in
Table 1 and is wrong. The parameters are named `fm_m2` / `fm_m3` (rather than
the `f_clform_<metab>` canonical used when a formation arm genuinely sits
*alongside* the elimination arm) precisely because the denominator each one
divides is the parent's whole clearance; see the `fm_<pathway>` entry in
`inst/references/parameter-names.md`.

# Units: a molar model validated against a mass-unit table

Leding 2026 fitted natural-log-transformed **molar** concentrations: Code S1
`$INPUT` declares `AMT ; dose in nmol` and `DV ; concentration in ln(nmol/L)`.
The packaged model therefore carries `dosing = "nmol"`, `concentration =
"nmol/L"`. But the dose *covariates* act on the mg dose (`DOSE ; dose in mg`),
which is why the model needs the separate `DOSE_TBAJ587_MG` column, and why
neither quantity can be derived from the other: **the paper reports no molecular
weight.**

That is not an obstacle for the parent, because the molecular weight cancels.
Writing `M` for the parent molecular weight, a `D` mg dose is `D * 1e6 / M` nmol,
concentrations scale linearly with the dosed amount, and converting back to mass
units multiplies by `M * 1e-6` -- so any mass-unit exposure is proportional to
`D` with `M` absent. Equivalently, dosing the numeric mg value into the nmol
slot returns parent concentrations directly in mg/L. The chunk below verifies
that invariance numerically rather than asserting it, then uses the resulting
scaffold throughout.

```{r mw-invariance}
mod <- rxode2::zeroRe(readModelDb("Leding_2026_tbaj587"))

# Helper: build a dosing + observation event table. Observation rows carry
# cmt = "central" (an ODE STATE, never the observable name "Cc") plus
# dvid = 1L to nominate an endpoint in this three-output model.
make_events <- function(dose_mg, dose_times, obs_times, fed, amt = dose_mg) {
  dos <- data.frame(time = dose_times, amt = amt, evid = 1L,
                    cmt = "depot", dvid = NA_integer_)
  obs <- data.frame(time = obs_times, amt = NA_real_, evid = 0L,
                    cmt = "central", dvid = 1L)
  ev <- rbind(dos, obs)
  ev <- ev[order(ev$time, ev$evid == 0L), ]
  ev$DOSE_TBAJ587_MG <- dose_mg
  ev$FED_HIGHFAT <- fed
  ev$id <- 1L
  ev
}

trapz <- function(x, y) sum(diff(x) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)

# Observation grid for the single-dose AUCinf work below. It must be FINE
# through absorption and distribution and may be coarse through the
# months-long terminal phase: a uniform 24 h grid understates AUC by ~0.3%
# because Tmax is around 4 h, which is enough to break a 1e-3 mass-balance
# gate for reasons that have nothing to do with the model.
auc_grid <- unique(c(seq(0, 24, by = 0.05),
                     seq(24, 312, by = 1),
                     seq(312, 24 * 365 * 8, by = 24)))

# Solve a single 200 mg fasted dose under two assumed molecular weights and
# convert each AUC back to mass units.
auc_mass <- vapply(c(500, 1000), function(MW) {
  ev <- make_events(200, 0, auc_grid, fed = 0, amt = 200 * 1e6 / MW)
  s <- as.data.frame(rxode2::rxSolve(mod, ev, atol = 1e-12, rtol = 1e-10,
                                     maxsteps = 1e6))
  s <- s[!is.na(s$Cc), ]
  trapz(s$time, s$Cc) * MW * 1e-6           # h*nmol/L -> h*mg/L
}, numeric(1))

# Deterministic (zeroRe) quantities, so this can be asserted tightly.
stopifnot(abs(auc_mass[1] / auc_mass[2] - 1) < 1e-6)
sprintf("AUCinf in h*mg/L at MW = 500 and 1000 g/mol: %.4f vs %.4f (ratio %.9f)",
        auc_mass[1], auc_mass[2], auc_mass[1] / auc_mass[2])
```

For the **metabolites** the cancellation is incomplete: a metabolite's mass-unit
exposure carries the ratio `MW_metabolite / MW_parent`. The paper states that
ratio only qualitatively, and only for M3 -- Discussion: "the difference in
molecular weight between TBAJ-587 and M3 is only 2%". The "Recovering the
molecular-weight ratios" section below turns that into a quantitative check.

Throughout the rest of this vignette the dose amount is the numeric mg value, so
`Cc` reads directly in mg/L and `Cc_m3` / `Cc_m2` read in *parent-molar-
equivalent* mg/L.

# Structural verification

## Stirling approximation in the absorption normaliser

Code S1 normalises the Savic gamma density with a Stirling approximation to
`log(NN!)` rather than the exact `lgamma(NN+1)`:

```
LNFAC = LOG(2.5066)+(NN+0.5)*LOG(NN)-NN
```

At `NN = 2.36` that understates `log(NN!)`, so the input rate integrates to
slightly **more** than `F * dose`. The model file reproduces the approximation
verbatim, because it is the expression against which every Table 1 value -- `CL`
above all -- was estimated. The factor is measured here so the mass-balance gate
below has an exact target instead of a fudge.

```{r stirling}
nn_hat  <- 2.36
lnfac   <- log(2.5066) + (nn_hat + 0.5) * log(nn_hat) - nn_hat
stirling_factor <- exp(lgamma(nn_hat + 1) - lnfac)
sprintf("Stirling LNFAC = %.6f vs exact lgamma(NN+1) = %.6f; input inflated by %.4f%%",
        lnfac, lgamma(nn_hat + 1), 100 * (stirling_factor - 1))
```

## Mass balance on all three analytes

This is the load-bearing gate. The `f(depot) <- 0` idiom that a NONMEM `F1 = 0`
requires can, for some models, make the whole `d/dt(depot)` right-hand side
evaluate to zero -- a flat-zero solve that produces **no error and no `NA`**,
and that ratio-based checks cannot detect because they divide one zero by
another. The only check that catches it is absolute mass balance, and it also
catches the opposite failure of a dose arriving twice.

At steady state after a single dose, `cl * AUCinf` must equal
`dose * F * stirling_factor` for the parent. For each metabolite, whose inflow is
`fm * cl * Cc` with `fm = 1` at the 200 mg fasted reference, `cl_metab *
AUCinf_metab` must equal the same quantity.

```{r mass-balance}
ev_mb <- make_events(200, 0, auc_grid, fed = 0)
s_mb <- as.data.frame(rxode2::rxSolve(mod, ev_mb, atol = 1e-12, rtol = 1e-10,
                                      maxsteps = 1e6))
s_mb <- s_mb[!is.na(s_mb$Cc), ]

# Apparent clearances at the 200 mg fasted reference: every dose factor is
# (200/200)^x = 1 and every fed factor is 1, so these are the Table 1 values.
target <- 200 * 1 * stirling_factor
mb <- tibble::tibble(
  Analyte = c("TBAJ-587", "M3", "M2"),
  `cl * AUCinf` = c(5.40 * trapz(s_mb$time, s_mb$Cc),
                    18.5 * trapz(s_mb$time, s_mb$Cc_m3),
                    34.1 * trapz(s_mb$time, s_mb$Cc_m2)),
  Target = target
) |>
  dplyr::mutate(Ratio = `cl * AUCinf` / Target)

# Deterministic; realised deviation is ~3e-5 on all three analytes, so 1e-3
# leaves room only for the trapezoid grid and Code S1's own 1e-5 log-guard
# offsets while still going red on a zeroed depot (ratio 0), a doubled dose
# (ratio 2), or a compensating parent loss term (metabolite ratios halve).
stopifnot(all(abs(mb$Ratio - 1) < 1e-3))

# Each analyte actually rises (the flat-zero failure mode this gate exists to
# catch), and nothing decays into negative solver noise over the 8-year
# horizon. Note Cc is exactly 0 at t = 0 by construction: f(depot) <- 0 means
# the dose never lands as a bolus, so `> 0` would fail on the pre-dose row.
stopifnot(max(s_mb$Cc) > 0, max(s_mb$Cc_m3) > 0, max(s_mb$Cc_m2) > 0)
stopifnot(all(s_mb$Cc >= 0), all(s_mb$Cc_m3 >= 0), all(s_mb$Cc_m2 >= 0))
knitr::kable(mb, digits = c(0, 4, 4, 6),
             caption = "Mass balance after a single 200 mg fasted dose.")
```

## The transit chain is live and `ktr = (NN+1)/MTT` is load-bearing

A model whose ODE chain had been silently discarded, or whose absorption input
had collapsed, would still reproduce AUC exactly -- AUC is `dose * F / CL`
whatever the absorption shape. Tmax is the discriminator. Quadrupling `MTT` on a
**single**-dose solve must move it materially; on a multiple-dose event table
Tmax would pin to the next dose time under any reading and the check would be
vacuous.

```{r mtt-sensitivity}
tmax_at <- function(mtt_factor) {
  m <- mod |> rxode2::ini(lmtt = log(0.837 * mtt_factor))
  ev <- make_events(200, 0, seq(0, 72, by = 0.05), fed = 0)
  s <- as.data.frame(rxode2::rxSolve(m, ev))
  s <- s[!is.na(s$Cc), ]
  s$time[which.max(s$Cc)]
}
tmax_1x <- tmax_at(1)
tmax_4x <- tmax_at(4)
stopifnot(tmax_4x - tmax_1x > 2)
sprintf("Tmax at MTT 0.837 h = %.2f h; at 3.348 h = %.2f h (shift %.2f h)",
        tmax_1x, tmax_4x, tmax_4x - tmax_1x)
```

# Replicating Table S2: AUC0-24h at the end of week 4

Table S2 is the paper's quantitative prediction table: `AUC0-24h` at the end of
week 4 of once-daily dosing in the fed state, without a loading period, at each
of the six studied dose levels, for all three analytes. It is simulated here
with typical values (`zeroRe`), which is what a "predicted" typical exposure
table calls for, and the interval AUC comes from **PKNCA** over the final
dosing interval.

```{r table-s2}
doses <- c(25, 50, 100, 200, 400, 800)
tau <- 24
start_ss <- 27 * tau        # start of the 28th (final) dosing interval
end_ss   <- 28 * tau

sim_s2 <- lapply(doses, function(D) {
  ev <- make_events(D, seq(0, start_ss, by = tau),
                    seq(start_ss, end_ss, by = 0.25), fed = 1)
  s <- as.data.frame(rxode2::rxSolve(mod, ev, atol = 1e-10, rtol = 1e-8,
                                     maxsteps = 1e6))
  s$treatment <- paste(D, "mg")
  s$amt_mg <- D
  s
}) |>
  dplyr::bind_rows() |>
  dplyr::filter(!is.na(Cc))

# One PKNCA pass per analyte (multi-output model). auclast over the final
# interval is AUC0-24h at the end of week 4.
auc_week4 <- function(conc_col) {
  nca_in <- sim_s2 |>
    dplyr::select(treatment, time, dplyr::all_of(conc_col)) |>
    dplyr::rename(Cc = dplyr::all_of(conc_col)) |>
    dplyr::mutate(id = 1L)

  dose_df <- sim_s2 |>
    dplyr::distinct(treatment, amt_mg) |>
    tidyr::crossing(time = seq(0, start_ss, by = tau)) |>
    dplyr::mutate(id = 1L, amt = amt_mg) |>
    dplyr::select(id, time, amt, treatment)

  conc_obj <- PKNCA::PKNCAconc(nca_in, Cc ~ time | treatment + id,
                               concu = "mg/L", timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                               doseu = "mg")
  intervals <- data.frame(start = start_ss, end = end_ss,
                          auclast = TRUE, cmax = TRUE, cmin = TRUE, cav = TRUE)
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                        intervals = intervals))
  as.data.frame(res$result)
}

nca_p  <- auc_week4("Cc")
nca_m3 <- auc_week4("Cc_m3")
nca_m2 <- auc_week4("Cc_m2")

nca_value <- function(df, param) {
  out <- df[df$PPTESTCD == param, c("treatment", "PPORRES")]
  # Guard against a silently-empty lookup (all(logical(0)) is TRUE).
  stopifnot(nrow(out) == length(doses))
  out$PPORRES[match(paste(doses, "mg"), out$treatment)]
}

published <- tibble::tibble(
  dose   = doses,
  p_pub  = c(2.08, 4.05, 7.83, 15.1, 28.8, 54.4),
  m3_pub = c(0.296, 0.532, 0.951, 1.69, 2.98, 5.21),
  m2_pub = c(0.139, 0.244, 0.425, 0.734, 1.25, 2.12)
)

cmp_s2 <- published |>
  dplyr::mutate(
    p_sim  = nca_value(nca_p,  "auclast"),
    m3_sim = nca_value(nca_m3, "auclast"),
    m2_sim = nca_value(nca_m2, "auclast"),
    p_pct  = 100 * (p_sim  - p_pub)  / p_pub,
    m3_pct = 100 * (m3_sim - m3_pub) / m3_pub,
    m2_pct = 100 * (m2_sim - m2_pub) / m2_pub
  )

cmp_s2 |>
  dplyr::select(dose, p_pub, p_sim, p_pct, m3_pub, m3_sim, m3_pct,
                m2_pub, m2_sim, m2_pct) |>
  dplyr::rename(
    "Dose (mg)"          = dose,
    "Parent pub"         = p_pub,
    "Parent sim"         = p_sim,
    "Parent %diff"       = p_pct,
    "M3 pub"             = m3_pub,
    "M3 sim"             = m3_sim,
    "M3 %diff"           = m3_pct,
    "M2 pub"             = m2_pub,
    "M2 sim"             = m2_sim,
    "M2 %diff"           = m2_pct
  ) |>
  knitr::kable(
    digits = 3,
    caption = paste(
      "AUC0-24h (h*mg/L) at the end of week 4, once-daily fed dosing, no",
      "loading period. Published values from Leding 2026 Table S2; simulated",
      "values are typical-value (zeroRe) PKNCA auclast over the final dosing",
      "interval. Metabolite simulated values are in parent-molar-equivalent",
      "mass units, so their offset is the molecular-weight ratio (see below)."
    )
  )
```

The **parent** reproduces Table S2 to a median absolute difference of
`r sprintf("%.2f%%", median(abs(cmp_s2$p_pct)))` across a 32-fold dose range.
That is a demanding joint test: it exercises the Stirling normaliser, the
`(NN+1)/MTT` transit rate, the power dose effect on clearance, both high-fat
meal effects, and 28 days of accumulation with a months-long terminal
half-life. A mis-transcribed clearance, dose reference or unit would move it by
tens of percent.

```{r table-s2-gate}
# Deterministic typical-value quantities, so these are tight by design.
stopifnot(max(abs(cmp_s2$p_pct)) < 1)
# Metabolites sit ~2.3% high purely from the molar-to-mass conversion; bound
# them well inside the 20% that would indicate a structural error.
stopifnot(max(abs(cmp_s2$m3_pct)) < 5, max(abs(cmp_s2$m2_pct)) < 5)
```

## Recovering the molecular-weight ratios

The metabolite columns are offset by a nearly constant factor across all six
dose levels. A *structural* error would not behave that way -- it would grow or
shrink with dose, because the dose covariates on `fm_m2`, `fm_m3` and `cl_m2`
are what make metabolite exposure dose-dependent. A constant multiplicative
offset is the signature of a unit conversion, and the only conversion left
unspecified is the molar-to-mass one.

Inverting it recovers each metabolite's molecular weight relative to the
parent's. For M3 the paper gives an independent check on the answer.

```{r mw-ratios}
mw_ratio <- tibble::tibble(
  Metabolite = c("M3", "M2"),
  `Implied MW ratio to parent` = c(median(cmp_s2$m3_pub / cmp_s2$m3_sim),
                                   median(cmp_s2$m2_pub / cmp_s2$m2_sim)),
  `Spread across the 6 doses` = c(
    diff(range(cmp_s2$m3_pub / cmp_s2$m3_sim)),
    diff(range(cmp_s2$m2_pub / cmp_s2$m2_sim))
  )
)
mw_ratio$`Implied MW difference` <-
  sprintf("%.1f%%", 100 * (1 - mw_ratio$`Implied MW ratio to parent`))

# The offset must be essentially dose-independent for the unit-conversion
# reading to hold at all; a structural error would not be.
stopifnot(all(mw_ratio$`Spread across the 6 doses` < 0.01))
# Leding 2026 Discussion: "the difference in molecular weight between TBAJ-587
# and M3 is only 2%". Recovered independently from Table S2.
stopifnot(abs(1 - mw_ratio$`Implied MW ratio to parent`[1]) < 0.04)

knitr::kable(mw_ratio, digits = 4, caption = paste(
  "Metabolite/parent molecular-weight ratios recovered by inverting the",
  "molar-to-mass conversion implied by Table S2. The M3 value is an",
  "independent check on the paper's stated 2% difference."
))
```

M3's recovered ratio of
`r sprintf("%.3f", mw_ratio[["Implied MW ratio to parent"]][1])` is a
`r sprintf("%.1f%%", 100 * (1 - mw_ratio[["Implied MW ratio to parent"]][1]))`
molecular-weight difference, matching the paper's stated "only 2%" and
confirming both that the metabolite offset is a unit conversion rather than a
transcription error and that the M3 sub-model is correct. M2's ratio is
recovered at
`r sprintf("%.3f", mw_ratio[["Implied MW ratio to parent"]][2])`; the paper
never states it, so this is a prediction rather than a confirmation.

# Dose proportionality and metabolite abundance

Results 3.2: "Increases in predicted exposure were approximately
dose-proportional for TBAJ-587, slightly less so for M2 and M3." Results 3.1:
"Abundance in plasma was in decreasing order: TBAJ-587, M3 and M2."

```{r dose-proportionality}
prop <- cmp_s2 |>
  dplyr::transmute(
    dose,
    `TBAJ-587` = (p_sim  / dose) / (p_sim[dose == 200]  / 200),
    M3         = (m3_sim / dose) / (m3_sim[dose == 200] / 200),
    M2         = (m2_sim / dose) / (m2_sim[dose == 200] / 200)
  )

# Parent is closer to proportional (normalised AUC flatter) than either
# metabolite: a deterministic ordering of three deterministic spreads.
span <- function(x) diff(range(x))
stopifnot(span(prop$`TBAJ-587`) < span(prop$M3),
          span(prop$`TBAJ-587`) < span(prop$M2))
# Parent stays within 25% of proportional across a 32-fold dose range.
stopifnot(span(prop$`TBAJ-587`) < 0.25)

# Abundance order, typical values at 200 mg fed.
ord <- cmp_s2[cmp_s2$dose == 200, ]
stopifnot(ord$p_sim > ord$m3_sim, ord$m3_sim > ord$m2_sim)

prop |>
  dplyr::rename("Dose (mg)" = dose) |>
  knitr::kable(digits = 3, caption = paste(
    "Dose-normalised AUC0-24h at the end of week 4, referenced to 200 mg.",
    "1.0 would be exact dose proportionality."
  ))
```

```{r figure-1-abundance}
sim_s2 |>
  dplyr::filter(treatment == "200 mg") |>
  dplyr::select(time, `TBAJ-587` = Cc, M3 = Cc_m3, M2 = Cc_m2) |>
  tidyr::pivot_longer(-time, names_to = "Analyte", values_to = "conc") |>
  ggplot(aes(time - start_ss, conc, colour = Analyte)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(x = "Time within the final dosing interval (h)",
       y = "Concentration (mg/L, metabolites parent-molar-equivalent)",
       title = "Analyte abundance at 200 mg once daily, end of week 4",
       caption = paste("Reproduces the ordering reported in Leding 2026",
                       "Results 3.1 and visible in Figure 1."))
```

# Terminal half-life across the dose range

Results 3.2: "The terminal half-life was predicted to range from 3.5 to
9.0 months for the explored dose range." Because apparent clearance rises with
dose, the shortest half-life belongs to the highest dose. This is an independent
check -- nothing in the transcription was fitted to it.

```{r half-life}
term_thalf <- function(D) {
  ev <- make_events(D, 0, seq(0, 24 * 365 * 3, by = 24), fed = 0)
  s <- as.data.frame(rxode2::rxSolve(mod, ev, atol = 1e-12, rtol = 1e-10,
                                     maxsteps = 1e6))
  s <- s[!is.na(s$Cc) & s$Cc > 0, ]
  # Fit the slope well after absorption and distribution have died away, so
  # this is the terminal phase and not a washout transient.
  tl <- s[s$time > 24 * 365 * 1.5, ]
  k <- -stats::coef(stats::lm(log(Cc) ~ time, data = tl))[["time"]]
  log(2) / k / (24 * 30.44)
}
thalf <- tibble::tibble(
  `Dose (mg)` = c(25, 800),
  `Terminal half-life (months)` = vapply(c(25, 800), term_thalf, numeric(1))
)

# Published range 3.5-9.0 months over 25-800 mg. Deterministic; 1 month of
# tolerance absorbs the choice of tail window without admitting a wrong model.
stopifnot(abs(thalf$`Terminal half-life (months)`[1] - 9.0) < 1,
          abs(thalf$`Terminal half-life (months)`[2] - 3.5) < 1)
stopifnot(thalf$`Terminal half-life (months)`[1] >
            thalf$`Terminal half-life (months)`[2])
knitr::kable(thalf, digits = 2, caption = paste(
  "Terminal half-life at the extremes of the studied dose range, single",
  "fasted dose. Leding 2026 Results 3.2 reports 3.5 to 9.0 months."
))
```

# Replicating Figure 3: loading doses and time to steady state

Leding 2026 explored 2-week loading periods at 2:1, 3:1 and 4:1 to the
maintenance dose, followed by 10 weeks of maintenance dosing, and compared them
against no loading period. `Cavg` is `AUC0-24h / 24` (Equation 5).

```{r figure-3}
build_regimen <- function(maint, ratio, weeks = 12) {
  ld_days <- if (ratio > 1) 14 else 0
  day <- seq(0, weeks * 7 - 1)
  dose_mg <- ifelse(day < ld_days, maint * ratio, maint)
  dos <- data.frame(time = day * 24, amt = dose_mg, evid = 1L, cmt = "depot",
                    dvid = NA_integer_, DOSE_TBAJ587_MG = dose_mg)
  ot <- seq(0, weeks * 7 * 24, by = 2)
  obs <- data.frame(
    time = ot, amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L,
    # The dose covariate is time-varying here: it takes the loading mg dose
    # during the loading period and the maintenance mg dose thereafter.
    DOSE_TBAJ587_MG = stats::approx(dos$time, dos$DOSE_TBAJ587_MG, ot,
                                    method = "constant", rule = 2)$y
  )
  ev <- rbind(dos, obs)
  ev <- ev[order(ev$time, ev$evid == 0L), ]
  ev$FED_HIGHFAT <- 1
  ev$id <- 1L
  ev$regimen <- if (ratio > 1) paste0(ratio, "x loading") else "no loading"
  ev
}

cavg_on_day <- function(s, day) {
  w <- s[s$time >= (day - 1) * 24 & s$time <= day * 24, ]
  trapz(w$time, w$Cc) / 24
}

regimens <- c(1, 2, 3, 4)
sim_load <- lapply(regimens, function(r) {
  ev <- build_regimen(100, r)
  s <- as.data.frame(rxode2::rxSolve(mod, ev, keep = "regimen",
                                     atol = 1e-10, rtol = 1e-8, maxsteps = 1e6))
  s <- s[!is.na(s$Cc), ]
  s$ratio <- r
  s
}) |>
  dplyr::bind_rows()

cavg_tab <- lapply(regimens, function(r) {
  s <- sim_load[sim_load$ratio == r, ]
  tibble::tibble(ratio = r,
                 regimen = if (r > 1) paste0(r, "x loading") else "no loading",
                 d14 = cavg_on_day(s, 14), d28 = cavg_on_day(s, 28),
                 d56 = cavg_on_day(s, 56), d84 = cavg_on_day(s, 84))
}) |>
  dplyr::bind_rows()

sim_load |>
  ggplot(aes(time / 24, Cc, colour = regimen)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = 14, linetype = "dotted") +
  labs(x = "Time (days)", y = "TBAJ-587 concentration (mg/L)",
       colour = "Regimen",
       title = "100 mg once daily, with and without a 2-week loading period",
       caption = paste("Replicates the 100 mg panel of Leding 2026 Figure 3.",
                       "Dotted line marks the end of the loading period."))
```

```{r figure-3-claims}
none <- cavg_tab[cavg_tab$ratio == 1, ]
load_only <- cavg_tab[cavg_tab$ratio > 1, ]

# 1. Loading doses raise early exposure (Results 3.3, "initially higher
#    concentrations"). Deterministic and large.
stopifnot(all(load_only$d14 / none$d14 > 1.5))

# 2. Steady state is NOT reached at 12 weeks: Cavg is still climbing between
#    week 8 and week 12 under every regimen.
stopifnot(all(cavg_tab$d84 / cavg_tab$d56 > 1.05))

# 3. The loading-vs-none gap narrows sharply once maintenance dosing starts.
#    This is the reproducible part of the paper's "similar at 4 weeks" claim;
#    see Assumptions and deviations for the part that is not.
gap_d14 <- load_only$d14 / none$d14 - 1
gap_d28 <- load_only$d28 / none$d28 - 1
stopifnot(all(gap_d28 < 0.5 * gap_d14))

cavg_tab |>
  dplyr::select(-ratio) |>
  dplyr::rename("Regimen" = regimen, "Day 14" = d14, "Day 28" = d28,
                "Day 56" = d56, "Day 84" = d84) |>
  knitr::kable(digits = 4, caption = paste(
    "Cavg (mg/L) of TBAJ-587 at 100 mg once daily maintenance, fed, by",
    "loading-dose ratio. Cavg = AUC0-24h / 24 (Leding 2026 Equation 5)."
  ))
```

# Between-subject variability

The simulations above are typical-value replications of the paper's own
typical-value tables. This section exercises the IIV structure: 12 etas, of
which 7 are fixed from the parent sub-model and 5 were estimated for the
metabolites.

```{r vpc-cohort}
# set.seed() seeds R's RNG, NOT rxode2's -- and rxode2's streams are
# partitioned per solver thread, so this cohort differs between a 16-thread
# workstation and a 2-core CI runner. No assertion below depends on the
# particular draw; see pattern 12 of known-vignette-failure-patterns.md.
set.seed(20260912)
n_per_arm <- 100L      # well under the 200/arm cap

vpc_arms <- c(50, 200, 800)
obs_grid <- c(seq(0, 24, by = 1), seq(30, 312, by = 6),
              seq(360, 126 * 24, by = 48))

events_vpc <- lapply(seq_along(vpc_arms), function(i) {
  D <- vpc_arms[i]
  base <- make_events(D, 0, obs_grid, fed = 0)
  base <- base[rep(seq_len(nrow(base)), times = n_per_arm), ]
  base$id <- rep((i - 1L) * n_per_arm + seq_len(n_per_arm),
                 each = nrow(base) / n_per_arm)
  base$treatment <- paste(D, "mg")
  base
}) |>
  dplyr::bind_rows()

# Disjoint ids across arms are mandatory: duplicates silently merge into one
# subject receiving the summed dose.
stopifnot(!anyDuplicated(unique(events_vpc[, c("id", "time", "evid")])))
stopifnot(dplyr::n_distinct(events_vpc$id) == n_per_arm * length(vpc_arms))

sim_vpc <- rxode2::rxSolve(readModelDb("Leding_2026_tbaj587"),
                           events = events_vpc, keep = "treatment") |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc))
```

```{r figure-s1-vpc}
sim_vpc |>
  dplyr::filter(time > 0, time <= 312) |>
  dplyr::group_by(treatment, time) |>
  dplyr::summarise(Q05 = quantile(Cc, 0.05), Q50 = quantile(Cc, 0.50),
                   Q95 = quantile(Cc, 0.95), .groups = "drop") |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~treatment) +
  scale_y_log10() +
  labs(x = "Time after a single dose (h)", y = "TBAJ-587 (mg/L)",
       title = "Simulated single-dose profiles with between-subject variability",
       caption = paste("Median and 5th-95th percentiles,", n_per_arm,
                       "subjects per arm. Comparable to the prediction-corrected",
                       "VPC in Leding 2026 Figure S1."))
```

```{r vpc-checks}
# The cohort median should track the typical-value prediction closely, and the
# IIV magnitude should land near the published CV%. Both are asserted as
# magnitudes with headroom, never as a sign, an equality, or a bound taken
# from one run.
cv_obs <- sim_vpc |>
  dplyr::filter(treatment == "200 mg", time == 312) |>
  dplyr::summarise(cv = stats::sd(Cc) / mean(Cc)) |>
  dplyr::pull(cv)

# Table 1 puts IIV on parent CL at 42.2% and on Vc at 48.2%; a single
# concentration late in the profile aggregates several of these. A cohort of
# 100 estimates a CV this size to roughly +/- 10% relative, so bound the
# plausible band generously -- it still goes red if the omega block were
# dropped (CV -> 0) or inflated by an order of magnitude.
stopifnot(cv_obs > 0.10, cv_obs < 1.50)
sprintf("Observed CV of Cc at 312 h, 200 mg arm: %.1f%% (n = %d)",
        100 * cv_obs, n_per_arm)
```

# Verification summary

```{r claims, echo = FALSE}
claims <- tibble::tribble(
  ~Claim, ~Source, ~Achieved, ~Status,
  "Mass balance cl*AUCinf = dose*F*Stirling, all 3 analytes",
  "Structural gate",
  sprintf("max |ratio - 1| = %.2g", max(abs(mb$Ratio - 1))),
  "Pass",

  "Parent AUC0-24h at week 4, 6 dose levels",
  "Table S2",
  sprintf("median |diff| %.2f%%, max %.2f%%",
          median(abs(cmp_s2$p_pct)), max(abs(cmp_s2$p_pct))),
  "Pass",

  "M3 AUC0-24h at week 4 (parent-molar-equivalent)",
  "Table S2",
  sprintf("max |diff| %.2f%%, resolved as a MW ratio", max(abs(cmp_s2$m3_pct))),
  "Pass",

  "M2 AUC0-24h at week 4 (parent-molar-equivalent)",
  "Table S2",
  sprintf("max |diff| %.2f%%, resolved as a MW ratio", max(abs(cmp_s2$m2_pct))),
  "Pass",

  "MW difference between TBAJ-587 and M3 is 'only 2%'",
  "Discussion",
  sprintf("recovered %.1f%% from Table S2",
          100 * (1 - mw_ratio[["Implied MW ratio to parent"]][1])),
  "Pass",

  "Approximately dose-proportional for the parent, less so for M2 and M3",
  "Results 3.2",
  sprintf("dose-normalised span: parent %.2f, M3 %.2f, M2 %.2f",
          span(prop$`TBAJ-587`), span(prop$M3), span(prop$M2)),
  "Pass",

  "Abundance order TBAJ-587 > M3 > M2",
  "Results 3.1",
  sprintf("%.2f > %.2f > %.2f h*mg/L", ord$p_sim, ord$m3_sim, ord$m2_sim),
  "Pass",

  "Terminal half-life 3.5 to 9.0 months over 25-800 mg",
  "Results 3.2",
  sprintf("%.2f months at 25 mg, %.2f at 800 mg",
          thalf$`Terminal half-life (months)`[1],
          thalf$`Terminal half-life (months)`[2]),
  "Pass",

  "Loading doses give initially higher concentrations",
  "Results 3.3",
  sprintf("day-14 Cavg +%.0f%% to +%.0f%% vs no loading",
          100 * min(gap_d14), 100 * max(gap_d14)),
  "Pass",

  "Steady state not reached after 12 weeks of dosing",
  "Results 3.3",
  sprintf("Cavg still rising: day84/day56 = %.2f to %.2f",
          min(cavg_tab$d84 / cavg_tab$d56), max(cavg_tab$d84 / cavg_tab$d56)),
  "Pass",

  "At 4 weeks, exposure similar between loading and no loading",
  "Results 3.3",
  sprintf("gap narrows from +%.0f%%..+%.0f%% (day 14) to +%.0f%%..+%.0f%% (day 28), not to ~0",
          100 * min(gap_d14), 100 * max(gap_d14),
          100 * min(gap_d28), 100 * max(gap_d28)),
  "Deviation"
)
knitr::kable(claims, caption = paste(
  "Published claims checked against the packaged model. Rows marked",
  "'Deviation' are documented below and excluded from the render gate."
))
```

```{r claims-gate}
stopifnot(all(claims$Status[claims$Status != "Deviation"] == "Pass"))
stopifnot(sum(claims$Status == "Deviation") == 1)
```

# Assumptions and deviations

* **Structure from Code S1, values from Table 1.** The supplement's final
  control stream carries the metabolite `$THETA` / `$OMEGA` records at their
  *initial* values, not their final ones (`CLM3` 20.121 vs 18.5; `Q3` 17.4191
  vs 9.09; `V6` 937.257 vs 1630). Every metabolite value in the model file
  therefore comes from Table 1, and every *equation form* from Code S1. The
  parent block is `FIX` in Code S1 at exactly its Table 1 values, so the two
  agree there; `e_dose_ka` is taken from Code S1 (`-0.000462`) because Table 1
  prints it rounded to two significant figures (`-0.00046`).

* **The Stirling approximation is reproduced, not corrected.** Code S1
  normalises the Savic density with
  `LNFAC = LOG(2.5066)+(NN+0.5)*LOG(NN)-NN` rather than the exact
  `lgamma(NN+1)`. At `NN = 2.36` that inflates the absorption input by
  `r sprintf("%.2f%%", 100 * (stirling_factor - 1))`, so the model absorbs
  `r sprintf("%.4f", stirling_factor)` times `F * dose` rather than exactly
  `F * dose`. Substituting the exact log-gamma would be "more correct" in
  isolation but would break consistency with the published `CL`, which was
  estimated against this expression; the same
  `r sprintf("%.2f%%", 100 * (stirling_factor - 1))` bias is embedded in
  Table 1. The mass-balance gate targets the inflated value explicitly so the
  approximation is visible rather than hidden.

* **Molar model, mass-unit validation table.** The model is in nmol and
  nmol/L, matching Code S1. The paper reports no molecular weights, so none is
  invented. Parent exposures in mass units are molecular-weight-independent
  (verified numerically above), so the Table S2 parent comparison is exact.
  Metabolite exposures are reported in *parent-molar-equivalent* mass units and
  carry the factor `MW_metabolite / MW_parent`, which this vignette recovers
  from Table S2 (M3 `r sprintf("%.3f", mw_ratio[["Implied MW ratio to parent"]][1])`,
  M2 `r sprintf("%.3f", mw_ratio[["Implied MW ratio to parent"]][2])`) rather
  than assuming. The recovered M3 ratio agrees with the paper's stated 2%
  difference; the M2 ratio is unstated in the paper and is a prediction.

* **`fm_m2` and `fm_m3` are anchored at 1, not summed to 1.** Both relative
  fractions metabolised are fixed at 1 and each metabolite receives the parent's
  whole elimination flux, so metabolite amounts are deliberately not
  mass-conserved against the parent and all metabolite clearances and volumes
  are apparent relative to an unidentifiable true fraction (Table 1 footnote a).
  This is faithful to Code S1, which retains the parent's full `-K20*A(2)` term.

* **`fixed()` marks only the by-fiat anchors.** `lfdepot`, `lfm_m2` and
  `lfm_m3` are wrapped in `fixed()`; they are the three parameters Table 1
  reports as "1 FIX" with no RSE and no confidence interval. Every other value,
  including `lnn`, carries an RSE and a SIR interval in Table 1 and so was
  estimated. Code S1 marks the whole parent block `FIX` because the final model
  estimated the metabolite sub-models with the parent held at its own final
  estimates -- an estimation-sequence device rather than a structurally fixed
  value, and not encoded as `fixed()` here.

* **Deviation: "at 4 weeks, the exposure was similar between no loading doses
  and loading doses" (Results 3.3) is not reproduced quantitatively.** At a
  100 mg maintenance dose the day-28 `Cavg` remains
  `r sprintf("+%.0f%%, +%.0f%% and +%.0f%%", 100 * gap_d28[1], 100 * gap_d28[2], 100 * gap_d28[3])`
  above the no-loading arm for the 2:1, 3:1 and 4:1 ratios. The *direction* of
  the claim is strongly reproduced -- the gap collapses from
  `r sprintf("+%.0f%%..+%.0f%%", 100 * min(gap_d14), 100 * max(gap_d14))` at
  day 14 to the day-28 values above -- and with a terminal half-life of months,
  drug delivered during a 2-week loading period cannot have washed out by week
  4, so a residual difference of this size is what the model's own structure
  requires. The likely reading is that the claim is a visual one about
  Figure 3, where the curves have converged relative to their loading-period
  separation on an axis spanning the full 12 weeks. The claim is reported as a
  deviation and excluded from the render gate rather than having the gate
  widened until it passed.

* **Not reproduced by design: the efficacy targets and safety references.**
  Efficacy Targets 1 and 2 (Equations 3 and 4) are derived from
  *bedaquiline* and bedaquiline-M2 typical daily average concentrations taken
  from a separate published model (Leding 2026 reference 27) together with
  MIC-ratio potency corrections, and the safety references come from a 3-month
  dog toxicity study (reference 26). Neither the upstream bedaquiline model
  predictions nor the dog toxicity exposures are reproducible from this paper
  alone, so Figures 3-5's horizontal target lines, and the conclusion that
  100 mg once daily is the lowest regimen meeting both efficacy targets, are
  outside the scope of this validation. The TBAJ-587 exposure predictions that
  those lines are compared *against* are validated above.

* **Illustrative cohort.** The between-subject-variability section simulates
  `r n_per_arm` subjects per arm at three of the six dose levels, fasted. The
  original observed data are not public, so it is a forward simulation from the
  packaged IIV rather than a true prediction-corrected VPC against Figure S1.
  Its assertions are written as magnitudes with headroom because rxode2's RNG
  streams are partitioned per solver thread and a CI runner draws a different
  cohort.

* **BQL handling is not part of the packaged model.** Leding 2026 omitted all
  parent below-quantification-limit records and all metabolite BQL records
  except the first and last in the decreasing and ascending curves, which were
  set to LLOQ/2 = 0.5 ng/mL (Results 3.1). That is an estimation-dataset
  decision with no simulation-time counterpart; simulated profiles here are not
  censored at the 1 ng/mL LLOQ.
