---
title: "Tramadol and O-desmethyltramadol (Lee 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tramadol and O-desmethyltramadol (Lee 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Lee_2019_tramadol"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
cat("* Article: <https://doi.org/10.2147/DDDT.S199574>\n", sep = "")
```

Lee et al. fitted tramadol and its active CYP2D6-derived metabolite
O-desmethyltramadol (M1) simultaneously in 22 healthy Korean male volunteers who
each took 100 mg of a sustained-release tramadol tablet every 12 h for five
doses. The paper is open access; there is no supplement and no NONMEM control
stream, so every value below is traced to the main text, Table 2 or Figure 2.

An erratum search (PubMed, the Dove Press article landing page and its
corrections feed, and a Google Scholar title search) returned no correction
notice for this article.

# Population

`r ui$population$n_subjects` of the 23 enrolled volunteers entered the final
model: one subject carrying CYP2D6\*5/\*5 was excluded before modelling
(Lee 2019 Methods, "Population PK model development"). All were Korean men aged
20-40 years (mean 24.8, SD 4.8), weighing 57-90 kg (mean 71.6, SD 8.9), 163-187
cm tall, with BMI 18.1-26.9 kg/m^2^ (Table 1). The genotype split among the 22
modelled subjects was 14 CYP2D6\*wt/\*wt and 8 CYP2D6\*10/\*10; no heterozygous
\*10 carrier was enrolled, and subjects with a \*5 allele or a duplicated CYP2D6
gene were excluded at screening.

Blood was drawn at 0, 0.5, 1, 1.5, 2, 2.5, 3, 4, 6, 8, 10, 12, 24, 48 and 72 h
after the fifth dose, yielding 328 tramadol and 323 M1 concentrations. The assay
was LC-MS/MS over 1-1000 ng/mL (tramadol) and 1-500 ng/mL (M1). Estimation was
by NONMEM 7.3 ADVAN6 with first-order conditional estimation and eta-epsilon
interaction.

The same information is available programmatically via
`readModelDb("Lee_2019_tramadol")()$population`.

# Model structure

Figure 2 of the paper draws the model explicitly, and the packaged file
reproduces it one arrow at a time:

* **Two parallel absorption inputs for one dose.** A fraction `Fr` = 0.88 enters
  `depot` and is absorbed first order at `ka` = 0.095 1/h (absorption half-life
  7.3 h, the sustained-release arm). The complementary 0.12 enters `central`
  directly as a zero-order input of duration `D2` = 1.93 h that starts after a
  lag `ALAG2` = 1.63 h. Lee 2019 Results: "We applied combined first- and
  zero-order absorption to catch the bimodal absorption phase of the extended
  release formulation of tramadol."
* **Two parallel exits from the tramadol compartment.** `k20` = CL/F / Vp/F is
  the non-M1 elimination route and `k23` = CLPM/F / Vp/F is the M1 formation
  route, so the total apparent tramadol clearance is CL/F + CLPM/F.
* **One metabolite compartment** eliminated by `k30` = CLM/F / Vm/F.

```{r structure, echo = FALSE}
knitr::kable(
  tibble::tribble(
    ~Item, ~Encoding,
    "ODE states", paste(ui$state, collapse = ", "),
    "Observations", paste(ui$predDf$var, collapse = ", "),
    "Analytic solution used?", ifelse(is.null(ui$linCmt), "no (explicit ODEs)", "YES - unexpected")
  ),
  caption = "Structure of the packaged model."
)
# The model defines a cl / vc pair, which rxode2 can silently replace with an
# analytic linCmt() solution, discarding the explicit d/dt lines and the
# metabolite compartment with them. Assert that it has not.
stopifnot(
  is.null(ui$linCmt),
  identical(ui$state, c("depot", "central", "central_m1")),
  identical(ui$predDf$var, c("Cc", "Cc_m1"))
)
```

## How the two absorption inputs are dosed

Each 100 mg administration is written as **two event records at the same time**
carrying the same `amt`: one bolus into `depot` and one zero-order input into
`central`. The `f(depot) <- ffo` / `f(central) <- 1 - ffo` lines in the model
apportion the dose between them, so the total delivered is one dose, not two.

The zero-order record is given an **explicit numeric `rate`**,
`amt * (1 - Fr) / D2`, rather than the usual `rate = -2` request for the
modelled duration. `rate = -2` currently fails in rxode2 5.1.8 for this
particular combination -- a two-endpoint model with both `dur()` and `alag()`
on the infused compartment, plus a second dose record at the same time -- with
`Duration is zero/negative`. The failure disappears if either endpoint is
removed or the `alag()` is dropped, so it is a solver interaction, not a model
defect. See "Assumptions and deviations".

The substitution is checked below rather than asserted in prose: on the
zero-order arm alone (which `rate = -2` does solve), the two encodings must
give the same profile to solver precision.

```{r dosing-helper}
# Fraction of the dose taking the first-order route; no IIV was reported on Fr,
# so this is a single number for every subject.
FFO <- plogis(as.numeric(ui$theta[["logitffo"]]))
stopifnot(abs(FFO - 0.88) < 1e-8)

# knitr::kable() emits markdown and pandoc reads a bare '*' as an emphasis
# marker, so "CYP2D6*10/*10" would render as "CYP2D610/10". Escape asterisks
# for DISPLAY only; the grouping labels themselves keep the published form.
md_esc <- function(x) gsub("*", "\\*", x, fixed = TRUE)

# Build an event table. `subj` must carry id, CYP2D6_STAR10_HOM and d1 (the
# subject's zero-order duration, needed to set the infusion rate).
make_events <- function(subj, n_doses, obs_times, tau = 12) {
  dose_times <- (seq_len(n_doses) - 1) * tau
  dose <- subj |>
    tidyr::crossing(time = dose_times) |>
    dplyr::mutate(amt = 100, evid = 1L, dvid = NA_integer_)
  dose <- dplyr::bind_rows(
    dose |> dplyr::mutate(cmt = "depot", rate = 0),
    dose |> dplyr::mutate(cmt = "central", rate = 100 * (1 - FFO) / d1)
  )
  obs <- subj |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(
      amt = NA_real_, evid = 0L, cmt = "central", rate = 0, dvid = 1L
    )
  dplyr::bind_rows(dose, obs) |>
    dplyr::select(-d1) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}
```

```{r rate-equivalence}
# Zero-order arm ONLY (no depot record), so rate = -2 solves. The numeric-rate
# substitution used everywhere else must reproduce it exactly.
d1_ref <- exp(as.numeric(ui$theta[["ld1"]]))
grid_eq <- seq(0, 48, by = 0.1)
ev_eq <- function(rate_value) {
  ev <- dplyr::bind_rows(
    data.frame(
      id = 1L, time = 0, amt = 100, evid = 1L, cmt = "central",
      rate = rate_value, dvid = NA_integer_
    ),
    data.frame(
      id = 1L, time = grid_eq, amt = NA_real_, evid = 0L, cmt = "central",
      rate = 0, dvid = 1L
    )
  )
  ev$CYP2D6_STAR10_HOM <- 0
  ev
}
mod_typ <- ui |> rxode2::zeroRe()
eq_modelled <- rxode2::rxSolve(mod_typ, ev_eq(-2), useLinCmt = FALSE) |>
  as.data.frame()
eq_numeric <- rxode2::rxSolve(
  mod_typ, ev_eq(100 * (1 - FFO) / d1_ref),
  useLinCmt = FALSE
) |>
  as.data.frame()
# Relative to the peak, because the residual is ODE-solver tolerance (rxode2
# defaults to rtol 1e-6), not a difference between the two encodings. A
# genuinely different infusion duration moves the profile by percent, so 1e-5
# separates "same solution" from "different model" by three orders of magnitude.
rate_agreement <- max(abs(eq_modelled$Cc - eq_numeric$Cc)) / max(eq_modelled$Cc)
stopifnot(nrow(eq_modelled) == length(grid_eq), rate_agreement < 1e-5)
```

The two encodings agree to `r signif(rate_agreement, 2)` of the peak
concentration, i.e. to the solver's own tolerance.

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Lee_2019_tramadol.R` carries an in-file comment
naming its source location; they are collected here for review.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` (CL/F) | 16.9 L/h | Table 2, row "CL/F (L/hr)" |
| `lvc` (Vp/F) | 59.9 L | Table 2, row "Vp/F (L)" |
| `lka` (ka) | 0.095 1/h | Table 2, row "ka (hr-1)" |
| `ld1` (D2) | 1.93 h | Table 2, row "D2 (hr)" |
| `ltlag` (ALAG2) | 1.63 h | Table 2, row "ALAG2 (hr)" |
| `logitffo` (Fr) | 0.88 | Table 2, row "Fr" |
| `lcl_form_m1` (CLPM/F) | 4.11 L/h | Table 2, row "CLPM/F (L/hr)" |
| `lcl_m1` (CLM/F) | 15.8 L/h | Table 2, row "CLM/F (L/hr)" |
| `lvc_m1` (Vm/F) | 8.63 L | Table 2, row "Vm/F (L)" |
| `e_cyp2d6_star10_hom_cl` | 0.351 | Table 2, row "CL/F, CYP2D6\*10/\*10"; Results equation CL/F = 16.9 (1 - 0.351 G) |
| `e_cyp2d6_star10_hom_cl_form_m1` | 0.528 | Table 2, row "CLPM/F, CYP2D6\*10/\*10"; Results equation CLPM/F = 4.11 (1 - 0.528 G) |
| `etalcl` | 0.059 | Table 2, row "omega^2 CL/F" (a variance) |
| `etalvc` | 0.023 | Table 2, row "omega^2 V/F" (a variance) |
| `etalcl_form_m1` | 0.017 | Table 2, row "omega^2 CLPM/F" (a variance) |
| `etald1` | 0.161 | Table 2, row "omega^2 D2" (a variance) |
| `propSd` | 0.13 | Table 2, row "sigma pro,tra"; Results "13%" |
| `addSd` | 2.99 ng/mL | Table 2, row "sigma add,tra"; Results "2.99 ng/mL" |
| `propSd_m1` | 0.109 | Table 2, row "sigma pro,ODMT"; Results "10.9%" |
| `addSd_m1` | 1.06 ng/mL | Table 2, row "sigma add,ODMT"; Results "1.06 ng/mL" |
| Parallel first- plus zero-order input, lagged zero-order arm | n/a | Figure 2 scheme (`ka`/`F1` into Depot, `D2`/`1-F1` into Tramadol); Results "Pharmacokinetic analysis" |
| Two parallel exits `k20` (CL) and `k23` (CLPM) from Tramadol | n/a | Figure 2 scheme |
| One-compartment M1 with first-order `k30` (CLM) | n/a | Figure 2 scheme; Results "Pharmacokinetic analysis" |
| Exponential IIV, combined proportional-plus-additive RUV | n/a | Methods "Population PK model development", unnumbered equations |
| Covariate form `theta_n1 * (1 - theta_n2 * G)` | n/a | Methods "Covariates selection", unnumbered equation |

# Typical-value verification of the covariate model

The paper states both covariate-adjusted clearances in words in the Discussion,
which makes them a direct, deterministic check on the encoding.

```{r typical-clearances}
typ_subj <- tibble::tibble(
  id = 1:2,
  genotype = c("CYP2D6*wt/*wt", "CYP2D6*10/*10"),
  CYP2D6_STAR10_HOM = c(0, 1),
  d1 = exp(as.numeric(ui$theta[["ld1"]]))
)

# A grid that is fine through absorption and the two zero-order kinks, and long
# enough for the tramadol terminal phase (absorption-rate limited, t1/2 7.3 h).
grid_sd <- sort(unique(c(
  seq(0, 24, by = 0.02), seq(24, 96, by = 0.25), seq(96, 480, by = 1)
)))

sim_typ <- rxode2::rxSolve(
  mod_typ,
  make_events(typ_subj, n_doses = 1L, obs_times = grid_sd),
  keep = "genotype",
  # rxode2's automatic ODE -> linCmt() conversion corrupts the dvid mapping for
  # multi-output models.
  useLinCmt = FALSE
) |>
  as.data.frame()

cl_tab <- sim_typ |>
  dplyr::group_by(genotype) |>
  dplyr::summarise(
    `CL/F (L/h)` = dplyr::first(cl),
    `CLPM/F (L/h)` = dplyr::first(cl_form_m1),
    `CLM/F (L/h)` = dplyr::first(cl_m1),
    `Total tramadol CL/F (L/h)` = dplyr::first(cl) + dplyr::first(cl_form_m1),
    .groups = "drop"
  )

knitr::kable(
  cl_tab |>
    dplyr::mutate(genotype = md_esc(genotype)) |>
    dplyr::rename("Genotype" = genotype),
  digits = 3,
  caption = "Typical apparent clearances by CYP2D6 star-10 genotype."
)

# Lee 2019 Discussion: "CL/F was estimated as 16.9 L/hr for wild type, and
# 11.0 L/hr for the CYP2D6*10/*10 group, while CLPM/F was estimated as 4.11
# L/hr and 1.94 L/hr for wild type and CYP2D6*10/*10 group, respectively."
# These are exact algebraic consequences of the ini() values, so the tolerance
# only has to absorb the paper's own rounding to 3 significant figures.
cl_wt <- cl_tab$`CL/F (L/h)`[cl_tab$genotype == "CYP2D6*wt/*wt"]
cl_hom <- cl_tab$`CL/F (L/h)`[cl_tab$genotype == "CYP2D6*10/*10"]
clpm_wt <- cl_tab$`CLPM/F (L/h)`[cl_tab$genotype == "CYP2D6*wt/*wt"]
clpm_hom <- cl_tab$`CLPM/F (L/h)`[cl_tab$genotype == "CYP2D6*10/*10"]
stopifnot(
  length(cl_wt) == 1L, length(cl_hom) == 1L,
  abs(cl_wt - 16.9) < 1e-8,
  abs(cl_hom - 11.0) < 0.05,
  abs(clpm_wt - 4.11) < 1e-8,
  abs(clpm_hom - 1.94) < 0.005
)
```

## The dose split and the zero-order duration are encoded as published

Two structural identities pin down the absorption model and the two parallel
exits. Both are exact for this linear model, so they are asserted tightly; a
mis-transcribed clearance, volume, dose split or unit moves them by percent, not
parts per million.

```{r closed-form-gates}
# Trapezoidal AUC on the (deterministic) typical-value profile.
trap_auc <- function(time, conc) {
  sum(diff(time) * (utils::head(conc, -1) + utils::tail(conc, -1)) / 2)
}

gates <- sim_typ |>
  dplyr::group_by(genotype) |>
  dplyr::summarise(
    auc = trap_auc(time, Cc),
    auc_m1 = trap_auc(time, Cc_m1),
    cl_tot = dplyr::first(cl) + dplyr::first(cl_form_m1),
    cl_form = dplyr::first(cl_form_m1),
    cl_m1 = dplyr::first(cl_m1),
    ctail = dplyr::last(Cc),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    # Whole dose recovered: (CL/F + CLPM/F) * AUCinf = Dose. 100 mg = 1e5 ng*L/mL.
    dose_recovery = cl_tot * auc / 1e5,
    # Formation flux recovered: CLM/F * AUCinf_M1 = CLPM/F * AUCinf_tramadol.
    m1_recovery = (cl_m1 * auc_m1) / (cl_form * auc)
  )

knitr::kable(
  gates |>
    dplyr::select(genotype, auc, auc_m1, dose_recovery, m1_recovery) |>
    dplyr::mutate(genotype = md_esc(genotype)) |>
    dplyr::rename(
      "Genotype" = genotype,
      "AUCinf tramadol (ng*h/mL)" = auc,
      "AUCinf M1 (ng*h/mL)" = auc_m1,
      "(CL+CLPM) * AUC / Dose" = dose_recovery,
      "CLM * AUC_M1 / (CLPM * AUC)" = m1_recovery
    ),
  digits = c(0, 1, 1, 6, 6),
  caption = "Closed-form mass-balance gates on the typical-value profile."
)

stopifnot(
  # The solve is deterministic (no random effects), so the only error here is
  # trapezoidal. Realised 3e-6 on this grid; 1e-3 leaves three orders of
  # magnitude of headroom and still breaks on any transcription error.
  all(abs(gates$dose_recovery - 1) < 1e-3),
  all(abs(gates$m1_recovery - 1) < 1e-3),
  # The profile has actually decayed, so AUCinf is not truncated.
  all(gates$ctail / (gates$auc / 480) < 1e-6)
)

# The zero-order arm really is zero-order, of duration D2, starting at ALAG2:
# dCc/dt must step UP by exactly the zero-order input rate at ALAG2 and step
# back DOWN by the same amount at ALAG2 + D2.
tlag_i <- exp(as.numeric(ui$theta[["ltlag"]]))
d1_i <- exp(as.numeric(ui$theta[["ld1"]]))
vc_i <- exp(as.numeric(ui$theta[["lvc"]]))
zo_rate_conc <- 1000 * 100 * (1 - FFO) / d1_i / vc_i

slope <- sim_typ |>
  dplyr::filter(genotype == "CYP2D6*wt/*wt", time <= 8) |>
  dplyr::mutate(dC = c(NA, diff(Cc) / diff(time)))
slope_at <- function(t) slope$dC[which.min(abs(slope$time - t))]

jump_up <- slope_at(tlag_i + 0.1) - slope_at(tlag_i - 0.1)
jump_down <- slope_at(tlag_i + d1_i - 0.1) - slope_at(tlag_i + d1_i + 0.1)

knitr::kable(
  tibble::tibble(
    "Zero-order kink" = c("start (ALAG2)", "end (ALAG2 + D2)"),
    "Time (h)" = c(tlag_i, tlag_i + d1_i),
    "Step in dCc/dt (ng/mL/h)" = c(jump_up, jump_down),
    "Expected (ng/mL/h)" = zo_rate_conc
  ),
  digits = 2,
  caption = "The zero-order arm delivers (1 - Fr) x Dose over D2, starting at ALAG2."
)

# The 0.2 h measurement window also sees the elimination term change, which is
# why the steps are not exact; realised 0.91 and 1.04 of the ideal step.
stopifnot(abs(jump_up / zo_rate_conc - 1) < 0.2, abs(jump_down / zo_rate_conc - 1) < 0.2)
```

# Virtual cohort

The observed data are not public, so the figures below use a virtual cohort of
200 subjects per genotype arm drawn from the published OMEGA.

The between-subject random effects are drawn **in R** with `stats::rnorm` and
passed to `rxSolve()` as per-subject typical values on a model whose random
effects have been zeroed. Two reasons: the subject's zero-order duration `D2`
has to be known before the event table can be written (it sets the infusion
rate, see above), and drawing in R makes the cohort reproducible across machines
and solver-thread counts, which `rxode2::rxSetSeed()` alone cannot guarantee.
The published OMEGA is diagonal, so independent normal draws reproduce it
exactly.

```{r cohort}
set.seed(20190601)

n_per_arm <- 200L
omega <- ui$omega
stopifnot(
  # Independent draws below are only equivalent to the published OMEGA because
  # the paper reports no covariances.
  isTRUE(all.equal(omega, diag(diag(omega)), check.attributes = FALSE)),
  identical(
    dimnames(omega)[[1]],
    c("etalcl", "etalvc", "etalcl_form_m1", "etald1")
  )
)

draw_arm <- function(n, genotype, star10_hom, id_offset) {
  tibble::tibble(
    id = id_offset + seq_len(n),
    genotype = genotype,
    CYP2D6_STAR10_HOM = star10_hom,
    etalcl = stats::rnorm(n, 0, sqrt(omega["etalcl", "etalcl"])),
    etalvc = stats::rnorm(n, 0, sqrt(omega["etalvc", "etalvc"])),
    etalcl_form_m1 = stats::rnorm(
      n, 0, sqrt(omega["etalcl_form_m1", "etalcl_form_m1"])
    ),
    etald1 = stats::rnorm(n, 0, sqrt(omega["etald1", "etald1"]))
  )
}

cohort <- dplyr::bind_rows(
  draw_arm(n_per_arm, "CYP2D6*wt/*wt", 0, 0L),
  draw_arm(n_per_arm, "CYP2D6*10/*10", 1, n_per_arm)
) |>
  dplyr::mutate(
    lcl = as.numeric(ui$theta[["lcl"]]) + etalcl,
    lvc = as.numeric(ui$theta[["lvc"]]) + etalvc,
    lcl_form_m1 = as.numeric(ui$theta[["lcl_form_m1"]]) + etalcl_form_m1,
    ld1 = as.numeric(ui$theta[["ld1"]]) + etald1,
    d1 = exp(ld1)
  )

cohort_params <- cohort |>
  dplyr::select(id, lcl, lvc, lcl_form_m1, ld1)

events <- make_events(
  cohort |> dplyr::select(id, genotype, CYP2D6_STAR10_HOM, d1),
  n_doses = 1L,
  obs_times = sort(unique(c(
    seq(0, 24, by = 0.2), seq(24, 96, by = 0.5), seq(96, 360, by = 2)
  )))
)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid", "cmt")])))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod_typ,
  events,
  params = cohort_params,
  keep = "genotype",
  useLinCmt = FALSE
) |>
  as.data.frame()

# The per-subject typical values really did reach the solver: each subject's
# cl must be exp(lcl_i) times its own genotype factor, and likewise for the
# formation clearance. This is the guard against rxSolve silently recycling
# the population typical value instead of the supplied per-subject row.
e_cl <- as.numeric(ui$theta[["e_cyp2d6_star10_hom_cl"]])
e_clpm <- as.numeric(ui$theta[["e_cyp2d6_star10_hom_cl_form_m1"]])

chk <- sim |>
  dplyr::group_by(id) |>
  dplyr::summarise(
    cl = dplyr::first(cl),
    cl_form_m1 = dplyr::first(cl_form_m1),
    .groups = "drop"
  ) |>
  dplyr::inner_join(
    cohort |> dplyr::select(id, lcl, lcl_form_m1, CYP2D6_STAR10_HOM),
    by = "id"
  ) |>
  dplyr::mutate(
    cl_expected = exp(lcl) * (1 - e_cl * CYP2D6_STAR10_HOM),
    cl_form_expected = exp(lcl_form_m1) * (1 - e_clpm * CYP2D6_STAR10_HOM)
  )
stopifnot(
  nrow(chk) == 2L * n_per_arm,
  # Both arms present, so a dropped covariate column would break this too.
  dplyr::n_distinct(chk$CYP2D6_STAR10_HOM) == 2L,
  max(abs(chk$cl - chk$cl_expected)) < 1e-8,
  max(abs(chk$cl_form_m1 - chk$cl_form_expected)) < 1e-8
)
```

## Simulated concentration-time profiles (replicates Figure 1)

Figure 1 of Lee 2019 shows the observed mean tramadol and M1 profiles for 72 h
after the last of five doses, stratified by CYP2D6 genotype, on a log
concentration axis. The single-dose cohort here reproduces the same
qualitative separation: tramadol exposure is higher and M1 exposure lower in
CYP2D6\*10/\*10 subjects.

```{r figure-1}
sim |>
  dplyr::filter(time <= 72) |>
  dplyr::select(id, genotype, time, Tramadol = Cc, `M1` = Cc_m1) |>
  tidyr::pivot_longer(c(Tramadol, `M1`), names_to = "analyte", values_to = "conc") |>
  dplyr::group_by(analyte, genotype, time) |>
  dplyr::summarise(
    Q05 = quantile(conc, 0.05),
    Q50 = quantile(conc, 0.50),
    Q95 = quantile(conc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50, colour = genotype, fill = genotype)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line() +
  facet_wrap(~analyte, scales = "free_y") +
  scale_y_log10() +
  labs(
    x = "Time after a single 100 mg dose (h)",
    y = "Concentration (ng/mL)",
    colour = NULL, fill = NULL,
    title = "Simulated single-dose profiles by CYP2D6*10 genotype",
    caption = "Median and 5th-95th percentile of 200 subjects per arm. Compare Figure 1 of Lee 2019."
  ) +
  theme(legend.position = "bottom")
```

# PKNCA validation

NCA is run separately for each analyte, with the genotype arm as the treatment
grouping variable.

```{r pknca-setup}
dose_df <- events |>
  dplyr::filter(evid == 1, cmt == "depot") |>
  dplyr::select(id, time, amt, genotype)
stopifnot(nrow(dose_df) == 2L * n_per_arm)

dose_obj <- PKNCA::PKNCAdose(
  dose_df, amt ~ time | genotype + id,
  doseu = "mg"
)

nca_intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE,
  half.life = TRUE, clast.obs = TRUE
)

run_nca <- function(conc_df, conc_col) {
  d <- conc_df
  d$Cc <- d[[conc_col]]
  d <- d |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::select(id, time, genotype, Cc)
  # Guarantee a time-zero anchor; pre-dose concentration is zero for an oral
  # single dose. Without it PKNCA warns once per subject that the AUC range
  # starts before the first measurement.
  d <- dplyr::bind_rows(
    d,
    d |> dplyr::distinct(id, genotype) |> dplyr::mutate(time = 0, Cc = 0)
  ) |>
    dplyr::distinct(id, genotype, time, .keep_all = TRUE) |>
    dplyr::arrange(id, time)
  stopifnot(nrow(d) > 0L, sum(d$time == 0) == 2L * n_per_arm)
  conc_obj <- PKNCA::PKNCAconc(
    d, Cc ~ time | genotype + id,
    concu = "ng/mL", timeu = "h"
  )
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = nca_intervals))
}

nca_tramadol <- run_nca(sim, "Cc")
nca_m1 <- run_nca(sim, "Cc_m1")
```

```{r pknca-table}
tidy_nca <- function(res, analyte) {
  as.data.frame(res) |>
    dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
    dplyr::group_by(genotype, PPTESTCD) |>
    dplyr::summarise(
      median = stats::median(PPORRES),
      p05 = stats::quantile(PPORRES, 0.05),
      p95 = stats::quantile(PPORRES, 0.95),
      .groups = "drop"
    ) |>
    dplyr::mutate(analyte = analyte)
}

nca_summary <- dplyr::bind_rows(
  tidy_nca(nca_tramadol, "Tramadol"),
  tidy_nca(nca_m1, "O-desmethyltramadol (M1)")
) |>
  dplyr::mutate(
    Parameter = dplyr::recode(
      PPTESTCD,
      cmax = "Cmax (ng/mL)",
      tmax = "Tmax (h)",
      aucinf.obs = "AUC0-inf obs (ng*h/mL)",
      half.life = "t1/2 (h)"
    )
  ) |>
  dplyr::select(analyte, genotype, Parameter, median, p05, p95) |>
  dplyr::arrange(analyte, Parameter, genotype)

nca_summary |>
  dplyr::mutate(genotype = md_esc(genotype)) |>
  dplyr::rename(
    "Analyte" = analyte,
    "Genotype" = genotype,
    "NCA parameter" = Parameter,
    "Median" = median,
    "5th pctile" = p05,
    "95th pctile" = p95
  ) |>
  knitr::kable(
    digits = 2,
    caption = paste(
      "Simulated single-dose NCA (PKNCA) after 100 mg oral tramadol,",
      "200 subjects per genotype arm."
    )
  )
```

All four terminal half-lives come out at about 7.3 h, and they are the same for
the parent and for the metabolite. That is the signature of flip-flop kinetics:
`ka` = 0.095 1/h gives an absorption half-life of 7.30 h, while the fastest
tramadol elimination half-life is 1.98 h (wild type) and the M1 elimination
half-life is only 0.38 h, so the absorption rate is what the terminal slope
measures in every arm and for both analytes. This is a structural consequence
of the published parameters, not a fitted quantity, so it is asserted directly.

```{r flipflop}
ka_halflife <- log(2) / exp(as.numeric(ui$theta[["lka"]]))
hl <- nca_summary |>
  dplyr::filter(Parameter == "t1/2 (h)") |>
  dplyr::mutate(rel = median / ka_halflife)
stopifnot(
  nrow(hl) == 4L,
  # Realised 1.005-1.009; lambda-z is fitted over a finite window, so it sits
  # slightly above the true asymptote. 3% still breaks if the terminal slope is
  # set by elimination (1.98 h wild type, 0.38 h for M1) instead of absorption.
  all(abs(hl$rel - 1) < 0.03)
)
```

Lee 2019 reports no NCA table, so there is nothing to place beside these values
with `ncaComparisonTable()`. The PKNCA output is instead checked against the
model's own closed-form exposure, which is the stronger test: for this linear
model each subject's AUC0-inf must equal `Dose / (CL/F + CLPM/F)` exactly, and
each subject's M1 AUC0-inf must equal `CLPM/F / CLM/F` times the tramadol
AUC0-inf.

```{r pknca-gates}
per_subject <- sim |>
  dplyr::group_by(id, genotype) |>
  dplyr::summarise(
    cl = dplyr::first(cl),
    cl_form_m1 = dplyr::first(cl_form_m1),
    cl_m1 = dplyr::first(cl_m1),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    auc_pred = 1e5 / (cl + cl_form_m1),
    auc_m1_pred = (cl_form_m1 / cl_m1) * auc_pred
  )

auc_obs <- function(res, nm) {
  d <- as.data.frame(res) |>
    dplyr::filter(PPTESTCD == "aucinf.obs") |>
    dplyr::select(id, PPORRES)
  names(d)[2] <- nm
  d
}

auc_cmp <- per_subject |>
  dplyr::inner_join(auc_obs(nca_tramadol, "auc_nca"), by = "id") |>
  dplyr::inner_join(auc_obs(nca_m1, "auc_m1_nca"), by = "id") |>
  dplyr::mutate(
    pct_diff = 100 * (auc_nca - auc_pred) / auc_pred,
    pct_diff_m1 = 100 * (auc_m1_nca - auc_m1_pred) / auc_m1_pred
  )
stopifnot(nrow(auc_cmp) == 2L * n_per_arm)

knitr::kable(
  auc_cmp |>
    dplyr::group_by(genotype) |>
    dplyr::summarise(
      `Tramadol median % diff` = stats::median(pct_diff),
      `Tramadol 90th pctile abs % diff` = stats::quantile(abs(pct_diff), 0.9),
      `M1 median % diff` = stats::median(pct_diff_m1),
      `M1 90th pctile abs % diff` = stats::quantile(abs(pct_diff_m1), 0.9),
      .groups = "drop"
    ) |>
    dplyr::mutate(genotype = md_esc(genotype)) |>
    dplyr::rename("Genotype" = genotype),
  digits = 3,
  caption = "PKNCA AUC0-inf against the model's closed-form AUC, per subject."
)

# The remaining difference is trapezoidal error plus PKNCA's extrapolation of
# the terminal tail, both of which depend on where each subject's zero-order
# kinks land on the observation grid. Assert the centre and a robust quantile
# rather than the extreme, and keep the bound loose enough to survive a
# different cohort draw but tight enough to break on a structural error (a
# wrong dose split, volume or clearance moves these by tens of percent).
stopifnot(
  abs(stats::median(auc_cmp$pct_diff)) < 1,
  stats::quantile(abs(auc_cmp$pct_diff), 0.9) < 2,
  abs(stats::median(auc_cmp$pct_diff_m1)) < 1,
  stats::quantile(abs(auc_cmp$pct_diff_m1), 0.9) < 2
)
```

# Steady state: replicating Figure 6

Figure 6 of Lee 2019 simulates tramadol and M1 "during 1 week after twice-daily
administration of 100 mg tramadol", and the Discussion draws the paper's
headline quantitative conclusion from it: "the peak plasma concentration of
tramadol for the CYP2D6 genotypes, CYP2D6\*10/\*10, was approximately 1.5 times
higher than that of the wild type (CYP2D6\*wt/\*wt), at the steady state after
multiple tramadol 100 mg twice daily administrations".

```{r steady-state}
tau <- 12
n_doses_ss <- 15L
last_full <- (n_doses_ss - 2L) * tau # start of the last COMPLETE interval

# Observe past the final dosing interval: rxode2 truncates a lagged infusion at
# the last observation record, which would silently clip the final interval.
grid_ss <- sort(unique(c(
  seq(0, last_full, by = 0.25),
  seq(last_full, last_full + tau, by = 0.02),
  seq(last_full + tau, (n_doses_ss - 1L) * tau + 2 * tau, by = 0.25)
)))

sim_ss <- rxode2::rxSolve(
  mod_typ,
  make_events(typ_subj, n_doses = n_doses_ss, obs_times = grid_ss, tau = tau),
  keep = "genotype",
  useLinCmt = FALSE
) |>
  as.data.frame()

sim_ss |>
  dplyr::filter(time <= 168) |>
  dplyr::select(genotype, time, Tramadol = Cc, `M1` = Cc_m1) |>
  tidyr::pivot_longer(c(Tramadol, `M1`), names_to = "analyte", values_to = "conc") |>
  ggplot(aes(time / 24, conc, colour = genotype)) +
  geom_line() +
  facet_wrap(~analyte, scales = "free_y") +
  labs(
    x = "Time (days)", y = "Concentration (ng/mL)", colour = NULL,
    title = "Typical-value profiles, 100 mg twice daily for one week",
    caption = "Replicates Figure 6 of Lee 2019."
  ) +
  theme(legend.position = "bottom")
```

```{r steady-state-metrics}
ss_metrics <- sim_ss |>
  dplyr::filter(time >= last_full, time <= last_full + tau) |>
  dplyr::group_by(genotype) |>
  dplyr::summarise(
    cmax = max(Cc),
    tmax = time[which.max(Cc)] - last_full,
    ctrough = dplyr::first(Cc),
    auctau = trap_auc(time, Cc),
    cmax_m1 = max(Cc_m1),
    auctau_m1 = trap_auc(time, Cc_m1),
    cl_tot = dplyr::first(cl) + dplyr::first(cl_form_m1),
    cl_form = dplyr::first(cl_form_m1),
    cl_m1 = dplyr::first(cl_m1),
    .groups = "drop"
  )

knitr::kable(
  ss_metrics |>
    dplyr::select(genotype, cmax, tmax, ctrough, auctau, cmax_m1, auctau_m1) |>
    dplyr::mutate(genotype = md_esc(genotype)) |>
    dplyr::rename(
      "Genotype" = genotype,
      "Tramadol Cmax,ss (ng/mL)" = cmax,
      "Tramadol Tmax,ss (h)" = tmax,
      "Tramadol Ctrough,ss (ng/mL)" = ctrough,
      "Tramadol AUCtau (ng*h/mL)" = auctau,
      "M1 Cmax,ss (ng/mL)" = cmax_m1,
      "M1 AUCtau (ng*h/mL)" = auctau_m1
    ),
  digits = 1,
  caption = "Typical-value steady-state exposure over the last complete 12 h interval."
)

wt <- ss_metrics[ss_metrics$genotype == "CYP2D6*wt/*wt", ]
hom <- ss_metrics[ss_metrics$genotype == "CYP2D6*10/*10", ]
stopifnot(nrow(wt) == 1L, nrow(hom) == 1L)

cmax_ratio <- hom$cmax / wt$cmax
m1_frac_ratio <- (hom$auctau_m1 / hom$auctau) / (wt$auctau_m1 / wt$auctau)

claims <- tibble::tribble(
  ~Claim, ~Source, ~Achieved, ~Pass,
  "Steady-state tramadol Cmax is about 1.5x higher in CYP2D6*10/*10",
  "Discussion", sprintf("%.2fx", cmax_ratio),
  cmax_ratio > 1.25 && cmax_ratio < 1.75,
  "AUCtau at steady state equals single-dose AUCinf (linear superposition)",
  "structural", sprintf("%.6f", hom$auctau / (1e5 / hom$cl_tot)),
  abs(hom$auctau / (1e5 / hom$cl_tot) - 1) < 1e-3,
  "M1-to-tramadol AUC ratio falls by the factor (1 - 0.528) in CYP2D6*10/*10",
  "Table 2 / Results", sprintf("%.5f", m1_frac_ratio),
  abs(m1_frac_ratio - (1 - 0.528)) < 1e-4
)

knitr::kable(
  claims |> dplyr::mutate(Claim = md_esc(Claim)),
  caption = "Published claims reproduced by the packaged model."
)
stopifnot(all(claims$Pass))
```

The M1-to-tramadol exposure ratio scales by exactly `1 - 0.528` between the two
genotype groups because CLM/F carries no genotype effect: the M1 formation
fraction is `CLPM/F / (CL/F + CLPM/F)`, but at steady state the M1 concentration
depends only on the formation flux `CLPM/F * Cc` and the M1 elimination
`CLM/F`, so the genotype effect on CLPM/F passes through undiluted. The
tramadol Cmax ratio (`r sprintf("%.2f", cmax_ratio)`) is smaller than the AUC
ratio (`r sprintf("%.2f", hom$auctau / wt$auctau)`) because the peak is still
partly absorption-limited, and it lands on the paper's stated "approximately
1.5 times".

# Assumptions and deviations

* **`omega^2 V/F` is carried on the parent volume.** Table 2 lists a single,
  unsubscripted "Variance of V/F" row while tabulating two volumes, `Vp/F`
  (tramadol) and `Vm/F` (M1). The Methods section names the parent volume
  "apparent volume of distribution (Vd/F)" and always writes the metabolite
  volume with its `m` subscript, so the eta is placed on the parent central
  volume. No IIV is carried on `Vm/F`, `CLM/F`, `ka`, `ALAG2` or `Fr`, matching
  the four IIV rows the paper reports.
* **The final covariate model has two covariate effects, not three.** Lee 2019
  Results ("Covariate analysis") states in prose that "the CL/F, `ka`, and the
  CLPM/F were significantly influenced" by CYP2D6\*10/\*10, but Table 2 lists
  covariate estimates only for CL/F and CLPM/F, Table 3's final model (row 4) is
  "Model 1 with CYP2D6\*10/\*10 as a covariate for CL/F, CLPM/F", and only two
  covariate equations are printed. The tables and equations are followed and the
  mention of `ka` is treated as a slip in the prose.
* **No molecular-weight correction on the parent-to-metabolite transfer.** The
  source ADVAN6 model moves amount from the tramadol compartment to the M1
  compartment through `k23` with no molar conversion (tramadol 263.4 g/mol
  versus M1 249.4 g/mol). That is reproduced verbatim. The consequence is that
  `Vm/F` and `CLM/F` are apparent values conditioned on the unknown
  bioavailability, on the unknown CYP2D6-mediated fraction of tramadol
  clearance, and on that 5% mass discrepancy; they should not be read as
  physiological M1 quantities. Their ratio, and therefore the M1 concentration
  the model predicts, is unaffected.
* **`CL/F` is the non-M1 elimination route, not total tramadol clearance.**
  Figure 2 draws `k20` (CL) and `k23` (CLPM) as two parallel arrows leaving the
  tramadol compartment, so the total apparent tramadol clearance in a wild-type
  subject is 16.9 + 4.11 = 21.0 L/h. Reading `CL/F` as the total would inflate
  every simulated concentration by about 24%.
* **Random effects are drawn in R, not by rxode2.** The subject's zero-order
  duration `D2` has to be known before the event table is written, because the
  zero-order arm is dosed with an explicit numeric rate (next bullet). Drawing
  the etas with `stats::rnorm` under `set.seed()` makes this vignette fully
  reproducible across machines and solver-thread counts. The published OMEGA is
  diagonal, so independent normal draws reproduce it exactly; this is asserted
  in the cohort chunk.
* **The zero-order arm is dosed with a numeric `rate` rather than `rate = -2`.**
  The model file keeps the published `dur(central) <- d1` and
  `alag(central) <- tlag`, which is the faithful encoding. rxode2 5.1.8,
  however, fails to solve a **two-endpoint** model that has both `dur()` and
  `alag()` on the infused compartment when a second dose record sits at the same
  time, raising `Duration is zero/negative`; the identical single-endpoint model
  solves fine, and so does the two-endpoint model once the `alag()` is removed.
  Supplying `rate = amt * (1 - Fr) / D2` bypasses the modelled-duration lookup
  and produces an identical solution (agreement to 3e-13 on the single-arm
  case). Users who dose this model with `rate = -2` and hit that error should
  use the numeric rate; nothing in the model file needs to change.
* **Observation records must extend past the final dosing interval.** rxode2
  truncates a lagged infusion at the last observation time, which silently clips
  about 10% off the final interval's AUC. The steady-state chunk therefore
  observes two dosing intervals beyond the analysis window and analyses the last
  *complete* interval.
* **The virtual cohort carries no demographic covariates.** Age, body weight,
  height and BMI were screened by Lee 2019 and none was retained, so none enters
  the model; they are recorded in the model file's `covariatesDataExcluded` for
  provenance only.
* **Single-dose simulation for the NCA block.** The study itself dosed five
  times every 12 h and sampled after the last dose. The NCA block uses a single
  dose because the closed-form gates it checks (`Dose / (CL/F + CLPM/F)` and the
  formation-flux identity) are cleanest without accumulation; the steady-state
  section covers the multiple-dose regimen, and both give the same AUC over a
  dosing interval, as asserted above.
* **No value in this vignette or the model file comes from anywhere other than
  the Lee 2019 main text, Table 2 or Figure 2.** There is no supplement, no
  control stream, no author correspondence and no figure digitisation.
