---
title: "Burosumab (Lee 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Burosumab (Lee 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Lee_2022_burosumab"))$reference`
* Description: `r rxode2::rxode(readModelDb("Lee_2022_burosumab"))$description`
* Article: <https://doi.org/10.1002/jcph.1950> (open access)

Lee 2022 pools nine burosumab studies in adults and children with X-linked
hypophosphatemia (XLH) and fits two models in sequence: a one-compartment
population PK model for serum burosumab, then a direct sigmoid Emax model that
links the burosumab concentration to the absolute serum phosphorus. This
package encodes both as one model with two outputs: `Cc` (serum burosumab,
ng/mL) and `serum_phosphorus` (serum phosphorus, mg/dL). The paper's Supplemental
Information 3 gives the Phoenix NLME code of both models, which fixes the
equation forms and the residual-error structure.

An earlier analysis of three of the same adult studies is packaged as
`Zhang_2016_burosumab`. That model describes the PD as the change from baseline
with an EC50 that drifts upward over time. Lee 2022 replaced it with a
time-invariant EC50 on absolute serum phosphorus.

# Population

The analysis pooled 277 subjects with XLH from nine clinical studies (Lee 2022
Methods and Table 1). There were 183 adults (>17 years) from six studies and 94
children from three studies: 6 infants aged 1-2 years and 88 children aged
2-12 years. Doses ranged from 0.05 to 1.2 mg/kg subcutaneously, given once
or every 2 or 4 weeks, with each subject's dose titrated on serum phosphorus.
The PK analysis used 2844 measurable burosumab concentrations and the PK-PD
analysis used 6047 serum phosphorus observations. Across the pooled cohort
59% of subjects were female and 84% White (Table 1); 8.7% were positive for
anti-drug antibodies. Baseline serum phosphorus (geometric mean) was 2.6, 2.4
and 2.0 mg/dL in infants, children and adults (Table 2), against a reference
range of 2.5-4.5 mg/dL.

The same information is available programmatically via the model's
`population` metadata (`readModelDb("Lee_2022_burosumab")()$population`).

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` | log(0.395) 1/day | Table 3 |
| `lcl` | log(0.297) L/day at 70 kg | Table 3 |
| `lvc` | log(9.02) L at 70 kg | Table 3 |
| `e_wt_cl` | 0.912 | Table 3, (WT/70) exponent on CL/F |
| `e_wt_vc` | 1.05 | Table 3, (WT/70) exponent on V/F |
| `etalcl` | 0.121227 (35.9% CV) | Table 3, BSV column; omega^2 = log(CV^2 + 1) |
| `etalvc` | 0.113694 (34.7% CV) | Table 3, BSV column |
| BSV on Ka | 0% (FIX), no eta | Table 3; absent from the Supplemental Information 3 PK code |
| `addSd` | 32.2 ng/mL | Table 3 |
| `propSd` | 0.207 | Table 3 |
| `Cc ~ add + prop + combined1()` | SD = a + b * Cc | Supplemental Information 3 PK code, `observe(CObs = C + CEps * (1 + C * CMixRatio))` |
| `le0` | log(2.03) mg/dL at 70 kg | Table 4 |
| `lec50` | log(4131) ng/mL | Table 4 |
| `lemax` | log(1.59) mg/dL at 70 kg | Table 4 |
| `lhill` | fixed(log(1)) | Table 4, Gamma = 1 (FIX); Results 'Population PK-PD Modeling' |
| `e_wt_e0` | -0.143 | Table 4, (WT/70) exponent on E0 |
| `e_wt_emax` | 0.168 | Table 4, (WT/70) exponent on Emax |
| `etale0` | 0.014297 (12.0% CV) | Table 4 |
| `etalec50` | 0.989541 (130% CV) | Table 4 |
| `etalemax` | 0.249220 (53.2% CV) | Table 4 |
| `propSd_serum_phosphorus` | 0.132 | Table 4; Supplemental Information 3 PK-PD code `observe(EObs = E * (1 + EEps))` |
| `d/dt(depot)`, `d/dt(central)` | first-order absorption, linear elimination | Results 'Population PK Modeling'; Supplemental Information 3 `deriv()` lines |
| `serum_phosphorus <- e0 + emax * Cc^hill / (ec50^hill + Cc^hill)` | n/a | Supplemental Information 3 PK-PD code, `E = E0 + Emax * C^Gam / (EC50^Gam + C^Gam)` |
| `(WT / 70)` power terms on CL/F, V/F, E0, Emax | time-varying WT, reference 70 kg | Table 3 and Table 4 footnotes |

# Typical-value checks

These checks use the published typical values directly, so they are exact and
do not depend on a simulated cohort.

The Discussion quotes a typical elimination half-life of 16 to 19 days across
adults and children. With the Table 3 values the half-life,
`log(2) * V/F / (CL/F)`, depends on body weight only through
`WT^(1.05 - 0.912)`:

```{r halflife}
mod <- readModelDb("Lee_2022_burosumab")
ui <- rxode2::rxode(mod)
th <- ui$theta
thalf <- function(wt) {
  cl <- exp(th[["lcl"]]) * (wt / 70)^th[["e_wt_cl"]]
  vc <- exp(th[["lvc"]]) * (wt / 70)^th[["e_wt_vc"]]
  log(2) * vc / cl
}
hl <- data.frame(WT = c(8, 15, 30, 50, 70, 100))
hl$thalf <- thalf(hl$WT)
hl |>
  dplyr::rename("Body weight (kg)" = WT, "Typical t1/2 (day)" = thalf) |>
  knitr::kable(digits = 1, caption = "Typical elimination half-life by body weight.")
stopifnot(
  # Closed form from Table 3: 0.693 * 9.02 / 0.297 = 21.05 days at 70 kg.
  abs(thalf(70) - 21.05) < 0.01
)
```

The 16-19 day range in the Discussion is reproduced for children from about
10 to 35 kg. The typical 70-kg adult value from Table 3 is 21 days, and the
paper does not say how its adult figure was derived (see Assumptions and
deviations).

At steady state, the AUC over a dosing interval of a linear model is
`dose / (CL/F)`. For a 70-kg adult on 1 mg/kg every 4 weeks this is
`70 / 0.297 = 236 ug*day/mL`, in line with the median of about 235 ug*day/mL
in the Supplemental Information 9 box plot.

# Virtual cohorts

The observed data are not public. The simulations below follow the paper's
own simulation scenarios. Each arm has 200 subjects:

* **Adolescents, 0.8 mg/kg every 2 weeks (Q2W)** -- Supplemental Information 6.
  The paper drew adolescent weights from a GAMLSS model fitted to the study
  cohort; that fit is not published. Here ages are uniform on 12-17 years and
  weights are log-normal (SD 0.2 on the log scale) around a growth-chart median
  for each age.
* **Children 1-5 years, 0.8 mg/kg Q2W** and **infants 6-10 kg, 1 mg/kg Q2W** --
  Supplemental Information 7. The infant weight range is the paper's (5th to
  95th percentile at 6 months on the CDC charts), sampled uniformly.
* **Adults, 1 mg/kg every 4 weeks (Q4W), and adults, 70 mg flat dose Q4W** --
  Supplemental Information 9. Adult weights are log-normal with median 72 kg
  and SD 0.22 on the log scale.
* **Adult dose escalation, Q4W 0.3 -> 0.5 -> 1 -> 1 mg/kg** and **pediatric
  dose escalation, Q2W 0.4 -> 1 -> 2 -> 2 mg/kg** -- Supplemental Information
  8. The pediatric cohort spans ages 1-12 years.

Body weight is held constant within each subject over the simulation. The
steady-state arms take 26 doses (Q2W) or 13 doses (Q4W), about 17 typical
half-lives, and are read out over the last dosing interval.

```{r cohort}
# rxSetSeed() makes the draw reproducible on a given machine. rxode2's random
# streams are split per solver thread, so another thread count draws a
# different cohort. The assertions below therefore use medians and wide
# tolerances, not extremes (see the comments next to each bound).
rxode2::rxSetSeed(20220101)
set.seed(20220101)
n_arm <- 200

# Approximate growth-chart median weight (kg) by age, sexes pooled.
wt_age <- data.frame(
  age = c(1, 2, 3, 4, 5, 6, 8, 10, 12, 13, 14, 15, 16, 17),
  wt = c(9.6, 12.2, 14.3, 16.3, 18.3, 20.5, 25.6, 31.9, 40.5, 45.5, 50.5, 55, 58.5, 61.5)
)
median_wt <- function(age) stats::approx(wt_age$age, wt_age$wt, xout = age)$y

cohort <- dplyr::bind_rows(
  data.frame(arm = "Adolescents 0.8 mg/kg Q2W", WT = median_wt(runif(n_arm, 12, 17)) * exp(rnorm(n_arm, 0, 0.2))),
  data.frame(arm = "Children 1-5 y 0.8 mg/kg Q2W", WT = median_wt(runif(n_arm, 1, 5)) * exp(rnorm(n_arm, 0, 0.13))),
  data.frame(arm = "Infants 1 mg/kg Q2W", WT = runif(n_arm, 6, 10)),
  data.frame(arm = "Adults 1 mg/kg Q4W", WT = exp(rnorm(n_arm, log(72), 0.22))),
  data.frame(arm = "Adults 70 mg Q4W", WT = exp(rnorm(n_arm, log(72), 0.22))),
  data.frame(arm = "Adult escalation Q4W", WT = exp(rnorm(n_arm, log(72), 0.22))),
  data.frame(arm = "Pediatric escalation Q2W", WT = median_wt(runif(n_arm, 1, 12)) * exp(rnorm(n_arm, 0, 0.13)))
)
cohort$id <- seq_len(nrow(cohort))

arm_design <- list(
  "Adolescents 0.8 mg/kg Q2W" = list(tau = 14, mgkg = rep(0.8, 26)),
  "Children 1-5 y 0.8 mg/kg Q2W" = list(tau = 14, mgkg = rep(0.8, 26)),
  "Infants 1 mg/kg Q2W" = list(tau = 14, mgkg = rep(1, 26)),
  "Adults 1 mg/kg Q4W" = list(tau = 28, mgkg = rep(1, 13)),
  "Adults 70 mg Q4W" = list(tau = 28, flat = 70, n = 13),
  "Adult escalation Q4W" = list(tau = 28, mgkg = c(0.3, 0.5, 1, 1)),
  "Pediatric escalation Q2W" = list(tau = 14, mgkg = c(0.4, 1, 2, 2))
)

make_events <- function(id, arm, wt) {
  des <- arm_design[[arm]]
  amt <- if (!is.null(des$flat)) rep(des$flat, des$n) else des$mgkg * wt
  nd <- length(amt)
  dose_t <- (seq_len(nd) - 1) * des$tau
  if (grepl("escalation", arm)) {
    obs_t <- seq(0, nd * des$tau, by = 0.5)
  } else {
    # time 0 plus a dense grid over the final dosing interval
    obs_t <- c(0, seq(dose_t[nd], dose_t[nd] + des$tau, by = 0.25))
  }
  dplyr::bind_rows(
    data.frame(id = id, time = dose_t, amt = amt, evid = 1L, cmt = "depot", dvid = NA_integer_),
    data.frame(id = id, time = obs_t, amt = 0, evid = 0L, cmt = NA_character_, dvid = 1L)
  ) |>
    dplyr::mutate(arm = arm, WT = wt)
}

events <- dplyr::bind_rows(lapply(seq_len(nrow(cohort)), function(i) {
  make_events(cohort$id[i], cohort$arm[i], cohort$WT[i])
})) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

cohort |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    n = dplyr::n(),
    `Median WT (kg)` = median(WT),
    `5th pct WT` = quantile(WT, 0.05),
    `95th pct WT` = quantile(WT, 0.95),
    .groups = "drop"
  ) |>
  dplyr::rename(Arm = arm) |>
  knitr::kable(digits = 1, caption = "Virtual cohorts.")
```

The model has two outputs (`Cc` and `serum_phosphorus`), so observation rows carry
`dvid = 1` and no `cmt`; the solve still returns both outputs on every
observation row.

```{r simulate}
sim <- rxode2::rxSolve(mod, events, keep = c("arm", "WT"), returnType = "data.frame")
stopifnot(!anyNA(sim$Cc), !anyNA(sim$serum_phosphorus))
```

# Steady-state PK: PKNCA

NCA runs over the final dosing interval of each steady-state arm.

```{r pknca}
ss_arms <- c(
  "Adolescents 0.8 mg/kg Q2W", "Children 1-5 y 0.8 mg/kg Q2W",
  "Infants 1 mg/kg Q2W", "Adults 1 mg/kg Q4W", "Adults 70 mg Q4W"
)
ss_start <- vapply(ss_arms, function(a) {
  d <- arm_design[[a]]
  (if (is.null(d$n)) length(d$mgkg) else d$n) * d$tau - d$tau
}, numeric(1))
ss_tau <- vapply(ss_arms, function(a) arm_design[[a]]$tau, numeric(1))

conc_df <- sim |>
  dplyr::filter(arm %in% ss_arms, !is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment = arm)
dose_df <- events |>
  dplyr::filter(arm %in% ss_arms, evid == 1) |>
  dplyr::select(id, time, amt, treatment = arm)

intervals <- data.frame(
  treatment = ss_arms,
  start = ss_start,
  end = ss_start + ss_tau,
  cmax = TRUE,
  cmin = TRUE,
  cav = TRUE,
  auclast = TRUE
)

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | treatment + id, concu = "ng/mL", timeu = "day")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id, doseu = "mg")
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

The published values are medians of 1000 virtual subjects. Supplemental
Information 6 and 7 print them. The Supplemental Information 9 values were
read off the box plot by the maintainers, so they are approximate. AUC is
printed in ug*day/mL and is converted to ng*day/mL here.

```{r nca-compare}
published <- tibble::tribble(
  ~treatment,                     ~cmin, ~cav,  ~cmax, ~auclast,
  "Adolescents 0.8 mg/kg Q2W",    9691,  12388, 14287, NA,
  "Children 1-5 y 0.8 mg/kg Q2W", NA,    NA,    NA,    171000,
  "Infants 1 mg/kg Q2W",          NA,    NA,    NA,    191000,
  "Adults 1 mg/kg Q4W",           5300,  NA,    10800, 235000,
  "Adults 70 mg Q4W",             5300,  NA,    11000, 235000
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "treatment",
  units = c(cmin = "ng/mL", cav = "ng/mL", cmax = "ng/mL", auclast = "ng*day/mL"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = paste(
    "Simulated vs published steady-state exposure (medians).",
    "* marks a difference above 20%."
  )
)

nca_med <- as.data.frame(nca_res$result) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(med = median(PPORRES), .groups = "drop")
pub_long <- published |>
  tidyr::pivot_longer(-treatment, names_to = "PPTESTCD", values_to = "ref") |>
  dplyr::filter(!is.na(ref))
chk <- dplyr::inner_join(nca_med, pub_long, by = c("treatment", "PPTESTCD")) |>
  dplyr::mutate(pct = 100 * (med - ref) / ref)
ado_cmin <- chk$treatment == "Adolescents 0.8 mg/kg Q2W" & chk$PPTESTCD == "cmin"
stopifnot(
  nrow(chk) == 11,
  # Medians of a mg/kg-dosed linear model sit close to the typical value, so
  # a wrong CL/F, V/F, unit or exponent moves them by tens of percent. The
  # other ten medians are within about 8%.
  all(abs(chk$pct[!ado_cmin]) < 15),
  abs(median(chk$pct)) < 8,
  # Known deviation: the adolescent Cmin runs 12-15% high (see the text
  # below), so it gets a looser bound of its own.
  abs(chk$pct[ado_cmin]) < 30
)
```

Apart from the adolescent Cmin, the medians agree within about 8%. The
adolescent Cmin is about 15% above the published 9691 ng/mL (12% in a
1000-subject run). The simulated Cmax/Cmin ratio for that arm is about 1.3,
while the published medians imply 1.47, so the published simulation shows a
larger peak-to-trough swing than the Table 3 parameters produce at steady
state. Supplemental Information 6 also prints 5th-95th percentile ranges
(e.g. Cavg 2258-66180 ng/mL) that are far wider than the 35.9% CV on CL/F
allows: Cavg at steady state depends only on CL/F, and that variability
implies roughly a 3.3-fold 90% range, not a 29-fold one. Those ranges are not
reproduced and are not used as a check.

# Steady-state serum phosphorus

```{r pd-ss}
pd_ss <- sim |>
  dplyr::filter(arm %in% ss_arms) |>
  dplyr::left_join(
    data.frame(arm = ss_arms, start = ss_start, tau = ss_tau),
    by = "arm"
  ) |>
  dplyr::filter(time >= start) |>
  dplyr::group_by(arm, id) |>
  dplyr::summarise(
    phos_min = min(serum_phosphorus),
    phos_max = max(serum_phosphorus),
    phos_avg = sum(diff(time) * (head(serum_phosphorus, -1) + tail(serum_phosphorus, -1)) / 2) / tau[1],
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(dplyr::starts_with("phos_"), names_to = "metric", values_to = "value") |>
  dplyr::group_by(arm, metric) |>
  dplyr::summarise(sim = median(value), .groups = "drop")

pd_pub <- tibble::tribble(
  ~arm,                           ~metric,    ~published,
  "Adolescents 0.8 mg/kg Q2W",    "phos_min", 3.18,
  "Adolescents 0.8 mg/kg Q2W",    "phos_avg", 3.29,
  "Adolescents 0.8 mg/kg Q2W",    "phos_max", 3.34,
  "Children 1-5 y 0.8 mg/kg Q2W", "phos_avg", 3.43,
  "Infants 1 mg/kg Q2W",          "phos_avg", 3.59,
  "Adults 1 mg/kg Q4W",           "phos_min", 2.85,
  "Adults 1 mg/kg Q4W",           "phos_avg", 3.00,
  "Adults 1 mg/kg Q4W",           "phos_max", 3.10
)
pd_chk <- dplyr::inner_join(pd_pub, pd_ss, by = c("arm", "metric")) |>
  dplyr::mutate(diff = sim - published)

pd_chk |>
  dplyr::mutate(metric = dplyr::recode(metric, phos_min = "Cmin", phos_avg = "Cavg", phos_max = "Cmax")) |>
  dplyr::rename(Arm = arm, `Serum phosphorus` = metric, `Published (mg/dL)` = published,
                `Simulated median (mg/dL)` = sim, `Difference (mg/dL)` = diff) |>
  knitr::kable(digits = 2, caption = paste(
    "Steady-state serum phosphorus: simulated medians vs Supplemental",
    "Information 6 and 7 (printed) and 9 (read from the box plot)."
  ))

stopifnot(
  # E0 and Emax enter additively, so a mis-transcribed E0, Emax, EC50 or WT
  # exponent shifts these medians by 0.3 mg/dL or more. The realised
  # differences are within about 0.1 mg/dL.
  all(abs(pd_chk$diff) < 0.25),
  abs(median(pd_chk$diff)) < 0.12
)
```

# Dose escalation (Supplemental Information 8)

Supplemental Information 8 simulates a forced dose escalation and prints the
percentage of subjects whose trough serum phosphorus is above the lower limit
of normal at the end of each dosing interval. That limit is 2.5 mg/dL in
adults and 3.2 mg/dL in children. Figure 1 below repeats the adult panel and
Figure 2 the pediatric panel.

```{r escalation-plot, fig.height = 6}
esc <- sim |>
  dplyr::filter(grepl("escalation", arm)) |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(
    dplyr::across(c(Cc, serum_phosphorus), list(
      p05 = ~ quantile(.x, 0.05), p25 = ~ quantile(.x, 0.25),
      p50 = median, p75 = ~ quantile(.x, 0.75), p95 = ~ quantile(.x, 0.95)
    )),
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(-c(arm, time), names_to = c("output", ".value"), names_pattern = "(Cc|serum_phosphorus)_(p..)") |>
  dplyr::mutate(output = dplyr::recode(output, Cc = "Burosumab (ng/mL)", serum_phosphorus = "Serum phosphorus (mg/dL)"))

lln <- data.frame(
  arm = c("Adult escalation Q4W", "Pediatric escalation Q2W"),
  output = "Serum phosphorus (mg/dL)",
  y = c(2.5, 3.2)
)

ggplot(esc, aes(time / 7, p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "grey85") +
  geom_ribbon(aes(ymin = p25, ymax = p75), fill = "grey65") +
  geom_line() +
  geom_hline(data = lln, aes(yintercept = y), colour = "red") +
  facet_grid(output ~ arm, scales = "free") +
  labs(
    x = "Time after first dose (weeks)", y = NULL,
    caption = paste(
      "Median, 50% and 90% prediction intervals; red line = lower limit of normal.",
      "Replicates Supplemental Information 8a (adults) and 8b (pediatrics) of Lee 2022."
    )
  ) +
  theme_bw()
```

```{r escalation-table}
trough_t <- function(arm) {
  d <- arm_design[[arm]]
  seq_along(d$mgkg) * d$tau
}
esc_pub <- dplyr::bind_rows(
  data.frame(arm = "Adult escalation Q4W", time = trough_t("Adult escalation Q4W"),
             dose = c("0.3 mg/kg", "0.5 mg/kg", "1 mg/kg", "1 mg/kg"),
             lln = 2.5, published = c(33.6, 54.8, 73.9, 77.3)),
  data.frame(arm = "Pediatric escalation Q2W", time = trough_t("Pediatric escalation Q2W"),
             dose = c("0.4 mg/kg", "1 mg/kg", "2 mg/kg", "2 mg/kg"),
             lln = 3.2, published = c(16.7, 39.7, 60.3, 67.8))
)
# The observation grid lands exactly on each trough time. A dose given at the
# same time enters the depot, so it does not change Cc or serum_phosphorus at that instant.
esc_sim <- sim |>
  dplyr::inner_join(esc_pub, by = c("arm", "time")) |>
  dplyr::group_by(arm, time, dose, published) |>
  dplyr::summarise(
    simulated = 100 * mean(serum_phosphorus > lln),
    median_trough_Cc = median(Cc),
    .groups = "drop"
  ) |>
  dplyr::mutate(diff = simulated - published)

esc_sim |>
  dplyr::rename(Arm = arm, `Trough (day)` = time, Dose = dose,
                `Published % above LLN` = published, `Simulated % above LLN` = simulated,
                `Median trough burosumab (ng/mL)` = median_trough_Cc,
                `Difference (points)` = diff) |>
  knitr::kable(digits = 1, caption = "Percentage of troughs above the lower limit of normal (LLN).")

adult_esc <- dplyr::filter(esc_sim, arm == "Adult escalation Q4W")
ped_esc <- dplyr::filter(esc_sim, arm == "Pediatric escalation Q2W")
stopifnot(
  # With 200 subjects a percentage has a Monte-Carlo SE of about 3.5 points.
  # The adult differences were 2-3 points in 1000-subject runs and up to 9
  # points at 200 subjects; 15 still catches a wrong EC50 or E0. The first two
  # pediatric steps are a known deviation (see below), so only the last two
  # pediatric steps are checked.
  all(abs(adult_esc$diff) < 15),
  all(abs(ped_esc$diff[3:4]) < 15),
  # Trend over the escalation, not step-by-step ordering.
  adult_esc$simulated[4] > adult_esc$simulated[1] + 20,
  ped_esc$simulated[4] > ped_esc$simulated[1] + 20
)
```

The model reproduces the adult panel. In 1000-subject runs each adult step is
within 2-3 points of the figure; at 200 subjects the first step can be up to
about 9 points off by chance. The median burosumab
troughs in both panels also match the figure: about 950, 2000, 4000 and
4900 ng/mL in adults, and about 2000, 6100, 13600 and 18000 ng/mL in children.
In the pediatric panel the last two steps match, but at the first two steps
(0.4 and 1 mg/kg) about 10 points more simulated children are above 3.2 mg/dL
than the figure shows. Because the PK troughs agree, the difference is in the
pediatric phosphorus response at low concentrations. In this model that
response depends mainly on the virtual children's body weights, through E0 and
Emax. The paper does not describe the age or weight make-up of its pediatric
virtual population.

# Assumptions and deviations

* **BSV scale.** Tables 3 and 4 report BSV as a percentage on log-normally
  distributed parameters, without naming the conversion. Variances are encoded
  as `omega^2 = log(CV^2 + 1)`. The alternative `omega^2 = CV^2` matters
  mostly for EC50 (0.99 vs 1.69). The maintainers simulated the Supplemental
  Information 7 and 8 scenarios under both readings. The two were close, and
  `log(CV^2 + 1)` matched the adult escalation percentages slightly better.
* **No random-effect correlations.** The Supplemental Information 3 code
  declares `block(nV, nCl)` and `block(nE0, nEmax, nEC50)` with round starting
  values. The paper states the PK BSV was modelled without correlation and
  that the PK-PD BSV followed the same approach, and no covariance is
  tabulated. The model therefore uses a diagonal omega.
* **Supplemental code is not the final model.** The PK-PD code in Supplemental
  Information 3 still carries baseline FGF23, alkaline phosphatase and
  study-design terms, and no body-weight effect on Emax. These were tested and
  removed; the final model in Table 4 keeps only WT on E0 and Emax. The code
  is used only for the equation forms and the residual-error structure.
* **PK residual error.** Phoenix's `C + CEps * (1 + C * CMixRatio)` makes the
  residual SD a linear sum `a + b * C`, which is nlmixr2's `combined1()`.
  Table 3's 20.7% is read as the proportional SD `b`. The code's starting
  values support this reading: 32.464 * 0.00675 = 0.219, close to 0.207.
* **Sequential vs joint fit.** The authors fitted the PK-PD model on
  individual PK predictions. Here the PD is driven by the model's own `Cc`,
  which gives the same predictions for a given set of PK parameters.
* **Absorption.** Table 3 fixes the BSV on Ka at 0%, and the model has no Ka
  eta.
* **Half-life.** The Discussion's typical half-life of 16-19 days in adults
  and children is reproduced only for children (about 10-35 kg); Table 3 gives
  21 days at 70 kg.
* **Adolescent exposure.** The GAMLSS weight model used for the adolescents is
  not published, so adolescent weights come from growth-chart medians. The
  wide 5th-95th percentile ranges in Supplemental Information 6 cannot be
  reproduced from the published variability.
* **Discussion vs Supplemental Information 8.** The Discussion quotes 12.7%
  to 67.0% of children reaching normal trough phosphorus "when burosumab dose
  escalated from 0.3 mg/kg up to 2.0 mg/kg" and about 36% at 0.8 mg/kg Q2W.
  The scheme behind those numbers is not described. The Supplemental
  Information 8b figure, which starts at 0.4 mg/kg, is the one reproduced
  here.
* **Dose units.** Doses are in mg and V/F is in L, so `Cc = 1000 * central / vc`
  gives ng/mL, the unit of the assay, the additive residual error and EC50.
