---
title: "Oxacillin (Lee 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Oxacillin (Lee 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Lee_2026_oxacillin"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1128/aac.01777-25> (open access, CC BY 4.0)
* Supplement: Table S1 (adverse events), Table S2 (raw per-patient data),
  Table S3 (raw concentrations), served by EuropePMC under PMC13231926.

Oxacillin is the standard-of-care antistaphylococcal penicillin for
methicillin-susceptible *Staphylococcus aureus*, but before this study no
pharmacokinetic data existed in neonates and young infants. Lee 2026 is a
prospective phase 1 study in 22 infants at most 90 days old, each given a
25 mg/kg loading dose over 30 min followed immediately by a continuous
infusion. The paper's clinical question is whether the unbound concentration
stays above the MIC for the whole dosing exposure (`fT 100% > MIC`), which for
a continuous infusion reduces to whether the steady-state unbound concentration
exceeds the MIC.

# Population

```{r population, echo = FALSE}
pop <- ui$population
tibble::tibble(
  Field = names(pop),
  Value = vapply(pop, function(x) paste(
    if (!is.null(names(x))) paste0(names(x), " ", x, "%") else as.character(x),
    collapse = "; "), character(1))
) |>
  knitr::kable(caption = "Population metadata carried in the model file.")
```

Twenty-two infants completed the study (Lee 2026 Table 1). Gestational age at
birth ranged from 23.9 to 40.3 weeks (median 38) and postnatal age at
enrolment from 4 to 82 days (median 32). Weights, which Table 1 does not
tabulate, are in Supplemental Table S2 and span 0.605 to 7.025 kg with a
median of 3.29 kg. Enrolment was stratified into four cohorts by gestational
age (<32 vs >=32 weeks) and postnatal age (<14 vs >=14 days) targeting six
infants each; the preterm-and-youngest cohort recruited only one of six, which
is why the paper declines to draw conclusions about gestational age. Twenty-one
infants received 160 mg/kg/day by continuous infusion and one received
120 mg/kg/day. Of 89 plasma samples 10 were below the 2 mcg/mL limit of
quantification and were excluded, leaving 79 concentrations for the fit.

The same information is available programmatically via
`readModelDb("Lee_2026_oxacillin")()$population`.

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL for 3.4 kg, 36 d) | 1.01 L/h | Table 2, `CL (Theta1)`; bootstrap 95% CI 0.791-1.3 |
| `lvc` (V for 3.4 kg) | 1.87 L | Table 2, `V (Theta2)`; bootstrap 95% CI 1.07-2.61 |
| `e_pna_cl` | 0.433 | Table 2, `Effect of PNA on CL (Theta3)`; bootstrap 95% CI 0.133-0.709 |
| `e_wt_cl` | 0.75 (fixed) | Table 2 footnote b `(WT_Kg/3.4)^0.75`; Methods states CL "was allometrically scaled by weight" -- a-priori value, no estimate or interval reported |
| `e_wt_vc` | 1 (fixed) | Table 2 footnote b `Vd = 1.87 * WT_Kg/3.4`, i.e. linear in weight |
| `etalcl` | 0.568 read as a log-scale SD; variance 0.322624 | Table 2, `IIV CL (etaCL)`; scale resolved below, not stated in the paper |
| `fu` | 0.1 (fixed) | Methods, "Simulations and pharmacodynamic assessments": free concentrations "estimated at 10% of total concentrations as reported in the literature (9)" |
| `addSd`, `propSd` | 0 (fixed) | Methods declares "a combined additive and proportional within-subject error model"; no magnitude is reported anywhere in the paper or its supplements |
| `d/dt(central) <- -kel * central` | n/a | Results, "Pharmacokinetic model": one-compartment, first-order elimination, NONMEM ADVAN1 TRANS2 |
| Reference weight 3.4 kg, reference PNA 36 d | n/a | Table 2 footnote b, and Results, "a typical infant 36 days postnatal age and weighing 3.4 kg" |

# Reading the published between-subject variability

Table 2 prints `IIV CL (etaCL) = 0.568` with no unit and no
variance / SD / CV qualifier. Read as a NONMEM `$OMEGA` **variance** it implies
a log-scale SD of `sqrt(0.568) = 0.754`; read as a **standard deviation** it is
0.568 itself. The two readings differ by a third in spread and change every
target-attainment number in the paper, so the choice cannot be left implicit.
Two independent checks against the paper's own data both exclude the variance
reading.

## Check 1 -- individual clearances recovered from Supplemental Table S2

Under a continuous infusion the steady-state concentration is exactly
`rate / CL`, so each infant's clearance is recoverable from the supplement
without any fitting. The table below transcribes, for all 22 infants,
Supplemental Table S2's current weight, dose-calculating weight, postnatal age,
daily dose and the two on-infusion samples (the 8-16 h and 16-96 h windows,
both at or beyond the ~5 half-lives needed for steady state).

```{r s2-transcribe}
s2 <- tibble::tribble(
  ~id,     ~wt,  ~dwt, ~pna, ~dose,     ~c2,   ~t2,     ~c3,   ~t3,
  "Pt1",  1.605, 1.605,   82,   160,  62.536, 10.07,  35.859, 81.07,
  "Pt3",  4.155, 4.155,   39,   160,  10.667, 10.38,  12.390, 30.15,
  "Pt4",  3.500, 3.500,   34,   160,  16.698,  8.67,   7.895, 69.70,
  "Pt5",  1.130, 1.130,   36,   160,      NA,    NA,   7.415, 16.93,
  "Pt6",  2.310, 2.000,    4,   160,      NA,    NA,  85.019, 17.00,
  "Pt8",  3.520, 3.515,   79,   160,  30.035, 14.88,  11.513, 35.35,
  "Pt9",  1.780, 1.785,   29,   160,      NA,    NA,  13.754, 46.87,
  "Pt11", 4.755, 3.170,   82,   160,  15.809,  8.57,  15.014, 18.13,
  "Pt12", 2.105, 2.105,   52,   160,      NA,    NA,  18.174, 39.08,
  "Pt14", 2.450, 2.500,    7,   160,  27.473, 11.20,  24.390, 35.28,
  "Pt16", 3.530, 3.530,    4,   160, 109.124,  8.28, 112.986, 16.43,
  "Pt17", 3.260, 3.260,   14,   160,   6.243,  8.15,  42.613, 44.93,
  "Pt18", 5.345, 5.345,   47,   160,  15.529, 11.10,  13.216, 17.67,
  "Pt21", 3.290, 3.290,   22,   160,  19.628, 11.70,  11.836, 16.57,
  "Pt22", 3.290, 3.290,   11,   160,  49.128,  9.08,  26.448, 17.55,
  "Pt23", 2.880, 2.880,   11,   160,  10.452, 12.38,  14.164, 63.35,
  "Pt24", 3.390, 3.390,   42,   160,  15.892,  8.75,  14.924, 16.23,
  "Pt25", 0.605, 0.697,    5,   120,      NA,    NA,  34.128, 16.00,
  "Pt26", 2.975, 2.975,   18,   160,  94.543,  9.08,  28.456, 59.58,
  "Pt27", 3.545, 3.500,   31,   160,  13.469, 12.63,  12.254, 17.52,
  "Pt28", 4.480, 4.480,   32,   160,  26.004,  8.35,  25.228, 16.60,
  "Pt29", 7.025, 7.025,   69,   160,  24.357,  9.20,   8.317, 15.88
)
stopifnot(nrow(s2) == 22L)

# Published covariate model, hard-coded from Table 2 footnote b so that this
# check is INDEPENDENT of what the model file happens to contain.
cl_published <- function(wt, pna_days) 1.01 * (wt / 3.4)^0.75 * (pna_days / 36)^0.433

s2 <- s2 |>
  mutate(
    rate   = dose * dwt / 24,             # mg/h delivered by the continuous infusion
    clpred = cl_published(wt, pna),
    eta2   = log((rate / c2) / clpred),   # log(CL_observed / CL_predicted)
    eta3   = log((rate / c3) / clpred),
    etabar = rowMeans(cbind(eta2, eta3), na.rm = TRUE)
  )

spread <- c(
  `16-96 h sample`   = sd(s2$eta3, na.rm = TRUE),
  `8-16 h sample`    = sd(s2$eta2, na.rm = TRUE),
  `per-infant mean`  = sd(s2$etabar)
)
round(spread, 3)
```

Every one of those numbers is an **upper bound** on the between-subject SD,
because it also contains residual error, assay error and any departure from
steady state. The SD reading (0.568) sits just underneath the bound, exactly
as a variance component should. The variance reading (0.754) exceeds it: a
between-subject SD cannot be larger than the total observed spread it is a
component of.

```{r s2-gate}
bound <- min(spread)
# Deterministic -- no simulation, no RNG. Realised bound 0.571.
stopifnot(
  0.568 < bound,   # SD reading is admissible
  0.754 > bound    # variance reading is not
)
```

## Check 2 -- the slope of the Figure 3 target-attainment curves

For a continuous infusion the paper's own criterion reduces to
`Css_unbound > MIC`, and `Css_unbound` is log-normal with log-scale SD equal to
the clearance SD. So `qnorm(PTA)` is **linear in `log(MIC)` with slope
`-1 / SD`**. The slope does not depend on the level, which means it is immune
to the virtual population's weight distribution -- something the paper never
reports.

The values below are read off the red 160 mg/kg/day curves of Figure 3's
`fT >100%` sub-panels. Points at or above 99% are excluded: they are at the
resolution limit of the plot and carry no slope information.

```{r fig3-slope}
fig3_ci <- tibble::tribble(
  ~pnad, ~mic,   ~pta,
      7, 0.25, 100.0,
      7, 0.50, 100.0,
      7, 1.00,  99.0,
      7, 2.00,  86.0,
     14, 0.25,  99.0,
     14, 0.50,  99.0,
     14, 1.00,  97.0,
     14, 2.00,  72.0,
     28, 0.25, 100.0,
     28, 0.50, 100.0,
     28, 1.00,  90.0,
     28, 2.00,  56.0,
     90, 0.25,  99.7,
     90, 0.50,  96.5,
     90, 1.00,  75.0,
     90, 2.00,  28.0
)

slope_fit <- fig3_ci |>
  filter(pta > 5, pta < 99) |>
  group_by(pnad) |>
  filter(n() >= 2) |>
  summarise(
    n_points = n(),
    omega    = -1 / coef(lm(qnorm(pta / 100) ~ log(mic)))[["log(mic)"]],
    .groups  = "drop"
  )

slope_fit |>
  mutate(omega = round(omega, 3)) |>
  rename("Postnatal age (days)" = pnad, "Usable points" = n_points,
         "Implied log-scale SD" = omega) |>
  knitr::kable(caption = "Level-free estimate of the clearance SD from the slope of Figure 3's continuous-infusion curves.")
```

```{r fig3-slope-gate}
om_fig <- mean(slope_fit$omega)
# Deterministic -- these are transcribed figure readings, not a simulation.
# Realised: mean 0.575, per-stratum 0.534 / 0.613 / 0.579.
stopifnot(
  abs(om_fig - 0.568) < abs(om_fig - 0.754),  # figure is nearer the SD reading
  abs(om_fig - 0.568) < 0.10,                 # and close to it in absolute terms
  all(slope_fit$omega < 0.70)                 # no stratum approaches 0.754
)
round(c(figure_implied = om_fig, sd_reading = 0.568, variance_reading = 0.754), 3)
```

Both checks agree, so the model file encodes `etalcl ~ 0.568^2 = 0.322624`.

# Structural checks

These are deterministic: they compare the packaged model against the published
equations and the published claims, with no random draw involved.

```{r structural}
mod <- readModelDb("Lee_2026_oxacillin")

# --- 1. Steady-state identity -------------------------------------------
# For a constant infusion, Css must equal rate / CL exactly, where CL comes
# from the hard-coded published equation above. This catches a mis-transcribed
# 1.01, 0.75, 0.433, 3.4, 36 or 1.87.
grid <- expand.grid(wt = c(1.0, 2.5, 3.4, 5.0, 7.0), pna = c(5, 14, 36, 60, 90))
ev_ss <- purrr::pmap_dfr(
  list(seq_len(nrow(grid)), grid$wt, grid$pna),
  function(i, wt, pna) {
    R <- 160 * wt / 24
    bind_rows(
      data.frame(id = i, time = 0,  amt = R * 240, evid = 1L, cmt = "central", rate = R),
      data.frame(id = i, time = c(200, 220, 240), amt = NA_real_, evid = 0L,
                 cmt = "central", rate = NA_real_)
    ) |> mutate(WT = wt, PNA = pna / 30.4375)
  }
)
ss <- rxode2::rxSolve(rxode2::zeroRe(mod), ev_ss, keep = c("WT", "PNA"),
                      returnType = "data.frame") |>
  group_by(WT, PNA) |>
  summarise(css = mean(Cc), .groups = "drop") |>
  mutate(
    pna_days = PNA * 30.4375,
    expected = (160 * WT / 24) / cl_published(WT, pna_days),
    pct_diff = 100 * (css - expected) / expected
  )
max(abs(ss$pct_diff))
stopifnot(max(abs(ss$pct_diff)) < 0.1)   # deterministic; realised 0.00

# --- 2. The published 2.2-fold claim -------------------------------------
# Results: PNA is "associated with a 2.2-fold increase in CL (weight adjusted)
# from 14 to 90 days of age".
fold_14_90 <- cl_published(3.4, 90) / cl_published(3.4, 14)
round(fold_14_90, 3)
stopifnot(abs(fold_14_90 - 2.2) < 0.05)

# --- 3. Half-life against the range the paper quotes ---------------------
# Methods cites "a half-life ranging from 1.2 to 3 hours according to GA and
# PNA (21)" for prior neonatal literature. Evaluate the model's typical-value
# half-life over the covariates of the 22 enrolled infants.
thalf <- log(2) * (1.87 * s2$wt / 3.4) / cl_published(s2$wt, s2$pna)
round(range(thalf), 2)
stopifnot(min(thalf) <= 1.2, max(thalf) >= 3.0)  # model brackets the quoted range
```

The half-life for the reference infant (3.4 kg, 36 days) is
`r round(log(2) * 1.87 / 1.01, 2)` h; across the enrolled cohort the model
spans `r round(min(thalf), 2)`-`r round(max(thalf), 2)` h, bracketing the
1.2-3 h the paper quotes from the prior literature.

# Virtual cohort and simulation

Three cohorts are simulated. Weights are held at published reference or cohort
values rather than sampled, because the paper does not report the virtual
population's covariate distribution -- see Assumptions.

```{r cohort}
# rxSetSeed() fixes rxode2's stream per solver thread, not across thread
# counts, so the cohort below differs between machines. Every assertion in
# this vignette is either deterministic or written to hold for any draw.
rxode2::rxSetSeed(20260914)

TEND  <- 48   # continuous infusion runs 0.5 -> 48 h
TOBS  <- 60   # observation continues through washout
WT_REF  <- 3.4;  PNA_REF  <- 36    # Lee 2026 Results, the "typical infant"
WT_PTA  <- 3.29                    # Supplemental Table S2 cohort median weight
N_ARM   <- 200L                    # per-arm cap

# Study regimen: 25 mg/kg over 30 min, then a continuous infusion.
study_arm <- function(ids, wt, pna, mgkgday, tend = TEND, tobs = TOBS) {
  R <- mgkgday * wt / 24
  dos <- tidyr::crossing(id = ids, k = 1:2) |>
    mutate(time = if_else(k == 1L, 0, 0.5),
           amt  = if_else(k == 1L, 25 * wt, R * (tend - 0.5)),
           rate = if_else(k == 1L, 25 * wt / 0.5, R)) |>
    select(-k)
  obs <- tidyr::crossing(
    id = ids,
    time = sort(unique(c(seq(0, tobs, by = 0.25), 0.5, tend)))
  ) |>
    mutate(amt = NA_real_, rate = NA_real_)
  bind_rows(mutate(dos, evid = 1L), mutate(obs, evid = 0L)) |>
    mutate(cmt = "central", WT = wt, PNA = pna / 30.4375) |>
    arrange(id, time, desc(evid))
}

# (a) typical-value arms: reference infant plus the four Figure 3 age strata
typ_spec <- tibble::tibble(
  cohort = c("Reference infant (3.4 kg, 36 d)",
             "PNA 7 d", "PNA 14 d", "PNA 28 d", "PNA 90 d"),
  wt  = c(WT_REF, rep(WT_PTA, 4)),
  pna = c(PNA_REF, 7, 14, 28, 90)
) |> mutate(id = row_number())

ev_typ <- purrr::pmap_dfr(
  list(typ_spec$id, typ_spec$wt, typ_spec$pna, typ_spec$cohort),
  function(i, wt, pna, ch) study_arm(i, wt, pna, 160) |> mutate(cohort = ch)
)

# (b) stochastic cohort at the reference covariates, for the Figure 2 band
ev_vpc <- study_arm(seq_len(N_ARM), WT_REF, PNA_REF, 160) |>
  mutate(cohort = "Reference infant (3.4 kg, 36 d)")

stopifnot(!anyDuplicated(unique(ev_typ[, c("id", "time", "evid")])))
stopifnot(!anyDuplicated(unique(ev_vpc[, c("id", "time", "evid")])))
```

```{r simulate}
sim_typ <- rxode2::rxSolve(rxode2::zeroRe(mod), ev_typ,
                           keep = c("cohort", "WT"), returnType = "data.frame")
sim_vpc <- rxode2::rxSolve(mod, ev_vpc,
                           keep = c("cohort", "WT"), returnType = "data.frame")
if (is.null(sim_typ$id)) sim_typ$id <- 1L
if (is.null(sim_vpc$id)) sim_vpc$id <- 1L
c(typical_rows = nrow(sim_typ), vpc_rows = nrow(sim_vpc))
```

# Replicate Figure 2

Figure 2 (top right) plots the unbound concentration of the reference infant
under the study regimen, with a 90% interval. The loading dose lifts the
unbound concentration to about 4 mg/L within 30 min and it then settles onto
the steady-state plateau.

```{r figure-2}
band <- sim_vpc |>
  filter(time <= TEND) |>
  group_by(time) |>
  summarise(Q05 = quantile(Cu, 0.05), Q50 = quantile(Cu, 0.50),
            Q95 = quantile(Cu, 0.95), .groups = "drop")
typ_ref <- sim_typ |>
  filter(cohort == "Reference infant (3.4 kg, 36 d)", time <= TEND)

ggplot(band, aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "#4444aa") +
  geom_line(linewidth = 0.8, colour = "#22226e") +
  geom_line(data = typ_ref, aes(time, Cu), colour = "firebrick",
            linetype = "dashed", linewidth = 0.7) +
  geom_hline(yintercept = c(0.5, 1), linetype = "dotted") +
  coord_cartesian(ylim = c(0, 8)) +
  scale_x_continuous(breaks = seq(0, 48, 12)) +
  labs(x = "Time (h)", y = "Unbound oxacillin concentration (mg/L)",
       title = "Figure 2 (top right) -- 25 mg/kg load + 160 mg/kg/day continuous infusion",
       subtitle = "Solid: cohort median with 90% interval. Dashed: typical value. Dotted: MIC 0.5 and 1 mg/L.",
       caption = "Replicates Figure 2 of Lee 2026 for a 3.4 kg, 36-day-old infant.")
```

```{r figure-2-numbers}
peak_u <- max(typ_ref$Cu)
css_u  <- typ_ref$Cu[typ_ref$time == TEND]
# Digitised from Figure 2 (top right), calibrated on the MIC = 0.5 and
# MIC = 1 mg/L dotted reference lines that the panel itself draws.
fig2 <- c(peak = 4.00, plateau = 2.48)
round(c(model_peak = peak_u, figure_peak = fig2[["peak"]],
        model_plateau = css_u, figure_plateau = fig2[["plateau"]]), 3)
round(100 * (c(peak_u, css_u) - fig2) / fig2, 1)

# Deterministic (zeroRe). The peak is set by the volume alone and matches to
# 0.5%. The plateau is set by the clearance and runs 9.5% low against the
# figure -- discussed under Assumptions; not tuned away.
stopifnot(abs(peak_u - fig2[["peak"]]) / fig2[["peak"]] < 0.05)
```

The peak is reproduced to within 0.5%. It is a pure test of the volume, since
both the 25 mg/kg dose and `Vd = 1.87 * WT/3.4` scale linearly with weight, so
`Dose/Vd` is weight-free. The plateau is `Css = rate / CL` and comes out 9.5%
below the digitised figure; see Assumptions.

# Replicate Figure 3 -- probability of target attainment

Figure 3 gives PTA against MIC for four postnatal ages and four regimens. For
the continuous infusion the paper uses the steady-state concentration as the
`fT 100% > MIC` surrogate; for the intermittent regimens `fT 100% > MIC` means
the unbound concentration stays above the MIC across the whole dosing interval,
which is evaluated here on the last complete interval of a 48 h simulation.

```{r pta-cohort}
regimens <- tibble::tribble(
  ~regimen,                 ~mgkgday, ~tau,
  "160 mg/kg/day CI",            160, NA_real_,
  "100 mg/kg/day div q6h",       100, 6,
  "200 mg/kg/day div q4h",       200, 4,
  "200 mg/kg/day div q6h",       200, 6
)
pnads <- c(7, 14, 28, 90)
MICS  <- c(0.25, 0.5, 1, 2)

pta_arm <- function(regimen, mgkgday, tau, pna, wt, n, id_offset) {
  ids <- id_offset + seq_len(n)
  if (is.na(tau)) {
    win <- 6                                     # window scored at steady state
    ev <- study_arm(ids, wt, pna, mgkgday, tend = TEND, tobs = TEND)
  } else {
    win <- tau
    per <- mgkgday * tau / 24 * wt               # mg per dose
    dos <- tidyr::crossing(id = ids, time = seq(0, TEND - tau, by = tau)) |>
      mutate(amt = per, rate = per / 0.5, evid = 1L)
    obs <- tidyr::crossing(
      id = ids,
      time = sort(unique(c(seq(0, TEND, by = 1), seq(TEND - tau, TEND, by = 1 / 30))))
    ) |>
      mutate(amt = NA_real_, rate = NA_real_, evid = 0L)
    ev <- bind_rows(dos, obs) |>
      mutate(cmt = "central", WT = wt, PNA = pna / 30.4375) |>
      arrange(id, time, desc(evid))
  }
  ev |> mutate(regimen = regimen, pnad = pna, win_h = win)
}

arms <- tidyr::crossing(regimens, pna = pnads) |> mutate(k = row_number())
ev_pta <- purrr::pmap_dfr(arms, function(regimen, mgkgday, tau, pna, k)
  pta_arm(regimen, mgkgday, tau, pna, WT_PTA, N_ARM, (k - 1L) * N_ARM))
stopifnot(!anyDuplicated(unique(ev_pta[, c("id", "time", "evid")])))

rxode2::rxSetSeed(20260914)
sim_pta <- rxode2::rxSolve(mod, ev_pta, keep = c("regimen", "pnad", "win_h"),
                           returnType = "data.frame")

pta <- lapply(MICS, function(m) {
  sim_pta |>
    filter(time >= TEND - win_h) |>
    group_by(regimen, pnad, id) |>
    summarise(hit = all(Cu > m), .groups = "drop") |>
    group_by(regimen, pnad) |>
    summarise(pta = 100 * mean(hit), .groups = "drop") |>
    mutate(mic = m)
}) |> bind_rows()
```

## Continuous infusion -- gated against Figure 3

```{r pta-continuous}
cmp_ci <- pta |>
  filter(regimen == "160 mg/kg/day CI") |>
  select(pnad, mic, simulated = pta) |>
  left_join(select(fig3_ci, pnad, mic, figure = pta), by = c("pnad", "mic")) |>
  mutate(difference = simulated - figure) |>
  arrange(pnad, mic)

cmp_ci |>
  mutate(across(c(simulated, figure, difference), ~round(.x, 1))) |>
  rename("Postnatal age (days)" = pnad, "MIC (mg/L)" = mic,
         "Simulated PTA (%)" = simulated, "Figure 3 PTA (%)" = figure,
         "Difference (pp)" = difference) |>
  knitr::kable(caption = "160 mg/kg/day continuous infusion: simulated fT 100% > MIC against the digitised red curves of Lee 2026 Figure 3.")
```

```{r pta-continuous-gate}
# 200 subjects per arm gives a Monte Carlo SE of up to 3.5 percentage points
# near PTA = 50%, and digitising Figure 3 is worth a further ~2 pp, so a
# per-point bound much below ~8 pp would sit inside the noise. Realised at
# 2 / 4 / 8 / 16 threads: mean |difference| 3.0-4.6 pp, max 13 pp (always at
# MIC 2, where the paper's virtual cohort evidently used lower,
# age-appropriate weights than the single 3.29 kg median assumed here).
stopifnot(
  mean(abs(cmp_ci$difference)) < 10,
  max(abs(cmp_ci$difference)) < 20
)
```

For a continuous infusion the PTA has a closed form -- `Css_unbound` is
log-normal, so `PTA = pnorm(log(Css_typical / MIC) / SD)`. Checking the
simulation against it separates "did the cohort solve correctly" from "does the
model agree with the paper", and lets the published 90% claims be tested
exactly rather than through Monte Carlo noise.

```{r pta-analytic}
pta_analytic <- function(mic, pna_days, wt = WT_PTA, sd_cl = 0.568) {
  css_u <- 0.1 * (160 * wt / 24) / cl_published(wt, pna_days)
  100 * pnorm(log(css_u / mic) / sd_cl)
}
cmp_ci <- cmp_ci |>
  mutate(analytic = pta_analytic(mic, pnad),
         mc_error = simulated - analytic)

# Monte Carlo agreement. Realised max |mc_error| 5.0 pp at 2/4/8/16 threads.
# A mis-transcribed clearance or exponent moves PTA by tens of points, so 10
# still goes red on a real error.
max(abs(cmp_ci$mc_error))
stopifnot(max(abs(cmp_ci$mc_error)) < 10)

# The paper's three headline claims, evaluated on the closed form so the test
# is deterministic. Claim 3 is reported but NOT gated: it fails reproducibly
# under this vignette's single-weight cohort (see Assumptions), and widening
# the gate until it passed would hide that.
claim_row <- function(text, value, holds, gated) {
  tibble::tibble(Claim = text, `Binding PTA (%)` = round(value, 1),
                 Holds = holds, Gated = gated)
}
a05 <- cmp_ci$analytic[cmp_ci$mic == 0.5]
a1  <- cmp_ci$analytic[cmp_ci$mic == 1 & cmp_ci$pnad <= 28]
a2  <- cmp_ci$analytic[cmp_ci$mic == 2]
claims <- dplyr::bind_rows(
  claim_row("All ages reach 90% PTA at MIC 0.5", min(a05), min(a05) >= 90, TRUE),
  claim_row("PNA <= 28 d reaches 90% PTA at MIC 1", min(a1), min(a1) >= 90, TRUE),
  claim_row("No age reaches 90% PTA at MIC 2 (CLSI breakpoint)", max(a2), max(a2) < 90, FALSE)
)
knitr::kable(claims, caption = "Published conclusions for 160 mg/kg/day continuous infusion, evaluated on the closed-form PTA. The third row is a recorded deviation.")
stopifnot(all(claims$Holds[claims$Gated]))
```

## Intermittent regimens -- reproduced, with a recorded deviation

```{r pta-intermittent}
fig3_int <- tibble::tribble(
  ~regimen,                ~pnad, ~mic, ~figure,
  "100 mg/kg/day div q6h",     7, 0.25,   54.5,
  "100 mg/kg/day div q6h",     7, 0.50,   38.0,
  "100 mg/kg/day div q6h",     7, 1.00,   21.0,
  "100 mg/kg/day div q6h",     7, 2.00,    8.5,
  "200 mg/kg/day div q4h",     7, 0.25,   96.0,
  "200 mg/kg/day div q4h",     7, 0.50,   91.0,
  "200 mg/kg/day div q4h",     7, 1.00,   78.0,
  "200 mg/kg/day div q4h",     7, 2.00,   47.5,
  "200 mg/kg/day div q6h",     7, 0.25,   67.0,
  "200 mg/kg/day div q6h",     7, 0.50,   54.0,
  "200 mg/kg/day div q6h",     7, 1.00,   38.0,
  "200 mg/kg/day div q6h",     7, 2.00,   21.0,
  "100 mg/kg/day div q6h",    90, 0.25,    6.0,
  "100 mg/kg/day div q6h",    90, 0.50,    1.5,
  "200 mg/kg/day div q4h",    90, 0.25,   50.0,
  "200 mg/kg/day div q4h",    90, 0.50,   35.0,
  "200 mg/kg/day div q4h",    90, 1.00,   17.0,
  "200 mg/kg/day div q4h",    90, 2.00,    4.0,
  "200 mg/kg/day div q6h",    90, 0.25,   11.0,
  "200 mg/kg/day div q6h",    90, 0.50,    6.0
)

cmp_int <- pta |>
  filter(regimen != "160 mg/kg/day CI") |>
  select(regimen, pnad, mic, simulated = pta) |>
  inner_join(fig3_int, by = c("regimen", "pnad", "mic")) |>
  mutate(difference = simulated - figure) |>
  arrange(pnad, regimen, mic)

cmp_int |>
  mutate(across(c(simulated, figure, difference), ~round(.x, 1))) |>
  rename("Regimen" = regimen, "Postnatal age (days)" = pnad, "MIC (mg/L)" = mic,
         "Simulated PTA (%)" = simulated, "Figure 3 PTA (%)" = figure,
         "Difference (pp)" = difference) |>
  knitr::kable(caption = "Intermittent regimens: simulated fT 100% > MIC against digitised Lee 2026 Figure 3. The model is systematically more optimistic; see Assumptions.")
```

### Controlling for the unreported virtual-population weight

The differences above mix two things: a genuine model-versus-figure
disagreement, and the fact that a single 3.29 kg weight is used for every
postnatal age while the paper's virtual cohort clearly used age-appropriate
weights. Both PTA criteria have a closed form -- the steady-state
concentration and the steady-state trough are each strictly decreasing in
clearance -- so the weight can be removed as a confounder without any
simulation.

```{r pta-closedform}
# Critical clearance above which the criterion fails, then PTA = P(CL < crit).
pta_cf <- function(mgkgday, tau, pna, wt, mic, sd_cl = 0.568) {
  v <- 1.87 * wt / 3.4
  crit <- if (is.na(tau)) {
    0.1 * (mgkgday * wt / 24) / mic                     # continuous: Css > MIC
  } else {
    dose <- mgkgday * tau / 24 * wt                     # intermittent: trough > MIC
    v * log1p(0.1 * dose / (v * mic)) / tau
  }
  100 * pnorm(log(crit / cl_published(wt, pna)) / sd_cl)
}
# Reproduces the simulation it is standing in for.
stopifnot(max(abs(sapply(MICS, function(m) pta_cf(160, NA, 90, WT_PTA, m)) -
                  pta_analytic(MICS, 90))) < 1e-8)

# Weight implied by each panel's OWN continuous-infusion curve. This makes the
# continuous rows below circular by construction; the intermittent rows are
# then out-of-sample predictions at that weight.
implied_wt <- function(pna) {
  d <- filter(fig3_ci, pnad == pna, pta > 5, pta < 99.5)
  optimise(function(w) sum((sapply(d$mic, function(m) pta_cf(160, NA, pna, w, m)) - d$pta)^2),
           interval = c(0.5, 8))$minimum
}
wt_imp <- vapply(c(7, 90), implied_wt, numeric(1))
names(wt_imp) <- c("7", "90")
round(wt_imp, 2)

cf <- tidyr::crossing(
  tibble::tibble(regimen = c("160 mg/kg/day CI", "100 mg/kg/day div q6h",
                             "200 mg/kg/day div q4h", "200 mg/kg/day div q6h"),
                 mgkgday = c(160, 100, 200, 200), tau = c(NA, 6, 4, 6)),
  pnad = c(7, 90), mic = MICS
) |>
  mutate(closed_form = purrr::pmap_dbl(
    list(mgkgday, tau, pnad, mic),
    function(d, t, p, m) pta_cf(d, t, p, wt_imp[[as.character(p)]], m))) |>
  inner_join(bind_rows(select(fig3_ci, pnad, mic, figure = pta) |>
                         mutate(regimen = "160 mg/kg/day CI"),
                       select(fig3_int, regimen, pnad, mic, figure)),
             by = c("regimen", "pnad", "mic")) |>
  mutate(difference = closed_form - figure) |>
  arrange(pnad, regimen, mic)

cf |>
  mutate(across(c(closed_form, figure, difference), ~round(.x, 1))) |>
  rename("Regimen" = regimen, "Postnatal age (days)" = pnad, "MIC (mg/L)" = mic,
         "Closed-form PTA (%)" = closed_form, "Figure 3 PTA (%)" = figure,
         "Difference (pp)" = difference) |>
  select(-mgkgday, -tau) |>
  knitr::kable(caption = "Closed-form PTA at the weight each panel's own continuous curve implies. The continuous rows are circular by construction; the intermittent rows are not.")
```

```{r pta-closedform-gate}
by_reg <- cf |> group_by(regimen) |> summarise(max_abs = max(abs(difference)), .groups = "drop")
by_reg |>
  mutate(max_abs = round(max_abs, 1)) |>
  rename("Regimen" = regimen, "Largest |difference| (pp)" = max_abs) |>
  knitr::kable(caption = "Worst-case agreement with Figure 3 once the weight confounder is removed.")

# Deterministic -- closed form, no RNG anywhere in this chunk.
# Realised: continuous 1.0 pp, q4h 6.7 pp, 100 q6h 23.1 pp, 200 q6h 23.1 pp.
q4h_err <- by_reg$max_abs[by_reg$regimen == "200 mg/kg/day div q4h"]
q6h_err <- max(by_reg$max_abs[grepl("q6h", by_reg$regimen)])
stopifnot(
  by_reg$max_abs[by_reg$regimen == "160 mg/kg/day CI"] < 3,   # circular, so tight
  q4h_err < 10,                                               # genuine out-of-sample agreement
  q6h_err > 10                                                # the disagreement is real, not noise
)
```

Once the weight is controlled for, the 200 mg/kg/day q4h curve is reproduced to
within 6.7 percentage points at both ages -- an out-of-sample agreement, since
nothing about that regimen entered the weight back-solve. Both q6h curves
remain 11-23 points too optimistic. The pattern is specific enough to point at
its cause: each published q6h curve tracks a simulation carrying **half** the
nominal amount per dose.

```{r pta-halfdose}
half <- tibble::tribble(
  ~published,              ~pnad, ~as_simulated,
  "200 mg/kg/day div q6h",     7, 100,
  "100 mg/kg/day div q6h",     7,  50,
  "200 mg/kg/day div q6h",    90, 100,
  "100 mg/kg/day div q6h",    90,  50
) |>
  tidyr::crossing(mic = MICS) |>
  mutate(half_dose = purrr::pmap_dbl(list(as_simulated, pnad, mic),
           function(d, p, m) pta_cf(d, 6, p, wt_imp[[as.character(p)]], m))) |>
  inner_join(select(fig3_int, regimen, pnad, mic, figure),
             by = c("published" = "regimen", "pnad", "mic")) |>
  mutate(difference = half_dose - figure)

half |>
  mutate(across(c(half_dose, figure, difference), ~round(.x, 1))) |>
  rename("Published label" = published, "Postnatal age (days)" = pnad,
         "Simulated as (mg/kg/day)" = as_simulated, "MIC (mg/L)" = mic,
         "Half-dose PTA (%)" = half_dose, "Figure 3 PTA (%)" = figure,
         "Difference (pp)" = difference) |>
  knitr::kable(caption = "Each Figure 3 q6h curve re-simulated at half the nominal per-dose amount.")

# Deterministic. Halving the per-dose amount cuts the worst error from 23.1 pp
# to 9.4 pp; anything under the full-dose error would be suggestive, and this
# is a threefold reduction.
stopifnot(max(abs(half$difference)) < q6h_err / 2)
round(max(abs(half$difference)), 1)
```

This is an observation about the published figure, not a change to the model:
the model file encodes the regimens as the paper's Methods describe them.

```{r pta-intermittent-check}
# NOT a pass/fail gate on the magnitude -- the disagreement is real and
# reproducible, so it is recorded rather than tuned away (see Assumptions).
# What IS gated is the ordering the paper's conclusion rests on, which the
# model does reproduce.
ord <- pta |>
  mutate(reg = recode(regimen,
                      "160 mg/kg/day CI" = "ci", "200 mg/kg/day div q4h" = "q4h",
                      "200 mg/kg/day div q6h" = "q6h200", "100 mg/kg/day div q6h" = "q6h100")) |>
  select(reg, pnad, mic, pta) |>
  tidyr::pivot_wider(names_from = reg, values_from = pta)

# Averaged over all 16 age x MIC points, so a single noisy cell cannot flip
# the ordering. Realised margins across 2/4/8/16 threads: 25-27, 12-14 and
# 13-15 percentage points respectively.
m <- summarise(ord, across(c(ci, q4h, q6h200, q6h100), mean))
round(unlist(m), 1)
stopifnot(m$ci - m$q4h > 10, m$q4h - m$q6h200 > 5, m$q6h200 - m$q6h100 > 5)

# Where the clinical decision is actually made (MIC >= 1) continuous beats the
# best intermittent regimen at every age. Realised minimum margin 7 pp.
stopifnot(all((ord$ci - ord$q4h)[ord$mic >= 1] > 3))

# The paper's intermittent conclusion: 200 mg/kg/day divided q4h -- eight times
# the FDA-labelled dose -- still misses the 90% target at the CLSI breakpoint.
# Realised maximum 75.5%.
stopifnot(all(ord$q4h[ord$mic == 2] < 88))
round(mean(cmp_int$difference), 1)
```

At the single 3.29 kg weight this cohort uses, the model predicts intermittent
PTA on average `r round(mean(cmp_int$difference), 0)` percentage points above
Figure 3, but the ordering the paper's conclusion rests on is reproduced:
continuous infusion beats every intermittent regimen, q4h beats q6h at the same
daily dose, and 200 mg/kg/day q4h -- eight times the FDA-labelled dose -- still
misses the 90% target at the CLSI breakpoint. The closed-form analysis above
localises the residual disagreement to the two q6h curves.

# PKNCA validation

NCA is run on the typical-value profiles: the loading-dose peak over the first
4 h, the steady-state interval 24-48 h, and the terminal phase after the
infusion stops at 48 h.

```{r pknca}
sim_nca <- sim_typ |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, cohort) |>
  # Keep Cc >= 1e-6 * max(Cc) per subject after the peak (solver noise below); time zero is re-added below.
  dplyr::group_by(id) |>
  dplyr::filter(time <= time[which.max(Cc)] | Cc >= 1e-6 * max(Cc)) |>
  dplyr::ungroup()

# Guarantee a time = 0 record per subject; concentration before the loading
# infusion starts is zero.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, cohort) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, cohort, time, .keep_all = TRUE) |>
  dplyr::arrange(id, cohort, time)
stopifnot(nrow(sim_nca) > 0, all(sim_nca$Cc >= 0))

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | cohort + id,
                             concu = "ug/mL", timeu = "h")

dose_df <- ev_typ |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, cohort)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | cohort + id, doseu = "mg")

intervals <- data.frame(
  start     = c(0,  24, 48),
  end       = c(4,  48, 60),
  cmax      = c(TRUE,  FALSE, FALSE),
  tmax      = c(TRUE,  FALSE, FALSE),
  cav       = c(FALSE, TRUE,  FALSE),
  cmin      = c(FALSE, TRUE,  FALSE),
  auclast   = c(FALSE, TRUE,  FALSE),
  half.life = c(FALSE, FALSE, TRUE)
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_wide <- nca_res$result |>
  dplyr::select(cohort, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
nca_wide |>
  dplyr::mutate(dplyr::across(where(is.numeric), ~round(.x, 2))) |>
  knitr::kable(caption = "Typical-value NCA by cohort. cmax/tmax over 0-4 h (loading dose), cav/cmin/auclast over the 24-48 h steady-state interval, half-life over the 48-60 h washout.")
```

## Comparison against published values

The only published NCA-comparable numbers are the two levels Figure 2 draws for
the reference infant: the unbound peak after the loading dose and the unbound
steady-state plateau, both converted to total concentration by the model's
fixed 10% unbound fraction.

```{r nca-compare}
published <- tibble::tibble(
  cohort = "Reference infant (3.4 kg, 36 d)",
  cmax   = fig2[["peak"]]    / 0.1,   # 4.00 mg/L unbound -> 40.0 ug/mL total
  cav    = fig2[["plateau"]] / 0.1    # 2.48 mg/L unbound -> 24.8 ug/mL total
)

cmp_nca <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by        = "cohort",
  units     = c(cmax = "ug/mL", cav = "ug/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp_nca, caption = "Simulated vs. Figure 2 of Lee 2026. * differs from reference by >20%.")
attr(cmp_nca, "footnote")
```

```{r nca-gate}
# Deterministic (typical values, zeroRe). Realised: cmax -0.5%, cav -9.5%.
pd <- as.numeric(gsub("[^0-9.+-]", "", cmp_nca$`% diff`))
stopifnot(max(abs(pd)) < 20)
```

# Assumptions, deviations and errata

* **Scale of the published IIV.** Lee 2026 Table 2 prints `IIV CL (etaCL) =
  0.568` without saying whether it is a variance, a standard deviation or a CV.
  It is encoded as a **log-scale standard deviation** (`etalcl ~ 0.568^2`) on
  the strength of the two checks above: the individual clearances recoverable
  from Supplemental Table S2 bound the between-subject SD at 0.571, and the
  slope of the Figure 3 continuous-infusion curves implies 0.575. The variance
  reading (SD 0.754) is excluded by both.

* **Residual error is not reported.** Methods state that "a combined additive
  and proportional within-subject error model was chosen", but neither
  magnitude appears in Table 2, in the Results text, or in any of the three
  supplemental workbooks. Both `addSd` and `propSd` are therefore encoded as
  `fixed(0)` rather than invented. Simulations from this model reproduce
  individual predictions, not observations; a user who needs residual noise
  must supply their own.

* **Figure 2 plateau runs 9.5% high against the model.** Digitising the
  Figure 2 top-right panel against the MIC 0.5 and MIC 1 reference lines the
  panel itself draws gives an unbound peak of 4.00 mg/L and a plateau of
  2.48 mg/L. The model, evaluated at the 3.4 kg / 36-day covariates the caption
  names, gives 3.98 and 2.24. The peak agrees to 0.5%; the plateau is 9.5%
  low. Since the peak depends only on the volume and the plateau only on the
  clearance, the gap is confined to CL and is consistent with the plotted
  profile having used a slightly younger or heavier infant than the caption
  states (2.48 mg/L corresponds to about 28 days at 3.4 kg, or about 5 kg at 36
  days). Table 2's `CL = 1.01` is used as printed.

* **Figure 3's two q6h curves do not reproduce; the q4h and continuous curves
  do.** Once the unreported virtual-population weight is controlled for by
  back-solving it from each panel's own continuous-infusion curve, the closed
  form reproduces the 200 mg/kg/day q4h curve to within 6.7 percentage points
  at both 7 and 90 days -- an out-of-sample agreement, since that regimen plays
  no part in the back-solve. Both q6h curves stay 11-23 points too optimistic.
  Re-simulating each q6h curve at **half** its nominal per-dose amount
  (25 rather than 50 mg/kg for 200 mg/kg/day q6h, 12.5 rather than 25 for
  100 mg/kg/day q6h) cuts the worst error from 23.1 to 9.4 points, so the
  published q6h curves behave like a half-dose simulation. Neither the paper
  nor its supplements describe how the intermittent regimens were coded
  (per-dose amount, infusion duration, which interval was scored), so this is
  reported as an observation about the figure and nothing in the model file was
  changed: the regimens are encoded exactly as Methods describes them. The
  paper's qualitative conclusion -- intermittent dosing up to 200 mg/kg/day
  fails the target that continuous infusion meets -- reproduces either way and
  is gated above.

* **One published claim does not reproduce: PNA 7 days at MIC 2.** Lee 2026
  concludes that "no patient, on continuous or intermittent infusions, achieved
  the PD target at the CLSI breakpoint MIC of 2", and Figure 3 puts the 7-day
  continuous-infusion curve at 86% there. The closed-form PTA for this
  vignette's cohort is 92.5% -- just over the 90% line -- and the simulation
  gives 90.5-95.0% depending on solver thread count. Every other age is well
  under 90% (81.8, 64.8 and 30.5% analytically at 14, 28 and 90 days). The gap
  is the weight assumption: a 7-day-old is materially lighter than the 3.29 kg
  cohort median used here, and `Css` scales as `WT^0.25`, so the youngest
  stratum is the one where the single-weight simplification bites hardest.
  This row is reported in the claims table and deliberately left out of the
  gate rather than widened until it passes.

* **Virtual-population covariates are an assumption.** The paper simulated
  10,000 virtual patients but does not report their weight distribution. The
  target-attainment cohort here fixes weight at the Supplemental Table S2
  cohort median of 3.29 kg for every postnatal age. Back-solving the paper's
  own continuous curves suggests it used age-appropriate weights (roughly
  1.5 kg at 14 days rising to 2.8 kg at 90 days), which is why the simulated
  PTA runs a little high at the youngest ages and matches closely at 90 days.
  Only the level is affected; the slope-based reading of the clearance SD is
  immune to it.

* **Postnatal age unit.** The canonical `PNA` covariate column carries months.
  Lee 2026 works in days and normalises to 36 days, so `model()` converts with
  1 month = 30.4375 days before forming the age ratio. Supply `PNA` in months.

* **Unbound fraction.** `fu = 0.1` was not measured in this study; Lee 2026
  took it from the literature (reference 9) and its Discussion warns that
  protein binding in preterm infants is probably lower than in adults, so the
  true unbound concentrations -- and therefore the true target attainment --
  may be higher than either the paper or this vignette reports.

* **Screened but unused covariates.** Gestational age, postmenstrual age,
  length, body surface area, albumin, serum creatinine and baseline
  transaminases were all tested and rejected; they are recorded in
  `covariatesDataExcluded` with the paper's stated reason, and none has a
  published point estimate.

* **Table 1 sex count.** Table 1's total male count (15, 68%) disagrees with
  the sum of its own per-cohort entries (16) and with Supplemental Table S2
  (16 male, 6 female). The population metadata follows the raw data.
