---
title: "Leuprolide testosterone suppression in normal and prostate-cancer rats (Lee 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Leuprolide testosterone suppression in normal and prostate-cancer rats (Lee 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Lee 2018 fits one PK-PD structure to four separate experimental groups of a
six-group rat study, so the paper contributes four model files. All four share
the testosterone feedback-turnover PD model; they differ in the leuprolide input
(intravenous bolus, subcutaneous solution, or a three-section sustained-release
microsphere depot) and in the strain / disease state.

```{r models}
model_names <- c(
  "Lee_2018_leuprolide_iv_rat",
  "Lee_2018_leuprolide_sc_rat",
  "Lee_2018_leuprolide_sr_rat",
  "Lee_2018_leuprolide_sr_cancer_rat"
)
uis <- lapply(model_names, function(n) rxode2::rxode(readModelDb(n)))
names(uis) <- model_names

tibble::tibble(
  Model = model_names,
  Group = c(
    "Group 1 - Wistar, IV leuprolide solution",
    "Group 3 - Wistar, SC leuprolide solution",
    "Group 4 - Wistar, SC Lucrin sustained-release depot",
    "Group 6 - Iar:COP (prostate cancer), SC Lucrin sustained-release depot"
  ),
  Species = vapply(uis, function(u) u$population$species, character(1))
) |>
  knitr::kable(caption = "The four models this paper contributes.")
```

* Citation: `r uis[[1]]$reference`
* Article: <https://doi.org/10.3390/molecules23040909>
* PubMed Central: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6017563/>

The paper is open access; the 16 numbered display equations were read from the
publisher's MathML rather than from the PDF text layer, and Figures 2, 3 and 5
were read from the publisher's figure files at native resolution.

# Population

Adult male rats, `n` = 5 per group, in six groups (Methods 4.3 and Table 5).
Normal animals were Wistar; the prostate-cancer animals were Iar:COP
(Copenhagen) rats carrying spontaneous Dunning R-3327 prostate adenocarcinoma.
Each animal received a single dose of 0.1 mg/kg leuprolide acetate - as a
solution intravenously (Group 1) or subcutaneously (Group 3), or as a Lucrin
depot microsphere dispersion subcutaneously (Groups 4 and 6) - or the dispersion
vehicle alone (Groups 2 and 5, which supplied the drug-free testosterone
baseline). Blood was sampled pre-dose and at 0.25, 1, 2, 4 and 8 h and at 1, 2,
3, 4, 5, 6, 7, 11 and 14 days. Leuprolide was assayed by UPLC-MS/MS and
testosterone by a competitive ELISA linear over 0.041-10 ng/mL (Methods 4.4,
4.5).

The same information is available programmatically from each model's
`population` metadata, e.g.
`rxode2::rxode(readModelDb("Lee_2018_leuprolide_iv_rat"))$population`.

## Body weight and the administered amount

Body weight is **not reported anywhere in the paper**, so the 0.1 mg/kg dose
cannot be converted to an amount directly. The paper's own definition of
clearance (Methods 4.6: "The clearance (CL) was calculated as the dose of
leuprolide divided by AUC0-inf") back-solves it from Table 1, once per route:

```{r dose}
dose_iv <- 105.50 * 248.61 # AUC0-inf (h*ng/mL) x CL (mL/h), Table 1 IV column
dose_sc <- 53.33 * 514.46 # AUC0-inf (h*ng/mL) x CL/F (mL/h), Table 1 SC column
c(dose_iv_ng = dose_iv, dose_sc_ng = dose_sc, implied_bw_g_iv = dose_iv / 100, implied_bw_g_sc = dose_sc / 100)
```

The two routes are independent arithmetic on different table columns yet imply
body weights of 262 g and 274 g - 4.5% apart, and both squarely in the normal
range for an adult male rat. That agreement is the reason these amounts are used
below rather than an assumed weight. The depot groups received "0.1 mg/kg (as
leuprolide)" (Table 5) and are given the subcutaneous amount.

# Source trace

Per-parameter provenance is recorded as an in-file comment beside each `ini()`
entry in the four files under `inst/modeldb/specificDrugs/`. Collected here:

| Equation / parameter | Value(s) | Source location |
|---|---|---|
| `d/dt(depot)`, `d/dt(central)`, `Cc` (solution) | n/a | Equations (1), (2), (3) |
| `d/dt(depot)`, `d/dt(depot2)`, `d/dt(transit1)`, `d/dt(transit2)`, `d/dt(central)`, `Cc` (depot) | n/a | Equations (4), (5), (6), (8), (9), (10) |
| `frel + frel2 + frel3 == 1` | n/a | Equation (7) |
| `d/dt(TT)` baseline form | n/a | Equation (11) |
| `TT(0) <- rbase_tt` | 4.353 (Wistar), 4.094 (Iar:COP) | Equations (12), (13) |
| `d/dt(TT)` drug-effect form | n/a | Equation (14) |
| `d/dt(moderator1)` | n/a | Equation (15) |
| `ce` sigmoid Emax | n/a | Equation (16) |
| `lka` | 16.67 1/h | Table 1, SC (Group 3), row `k a` |
| `lcl` | 248.61 mL/h (IV); 514.46 mL/h (SC and both depots) | Table 1, row `CL` |
| `lvc` | 192.95 mL (IV); 487.40 mL (SC and both depots) | Table 1, row `V d` |
| `lka2` (paper's `kd`) | 0.08 1/h both strains | Table 2, row `k d` |
| `lktr` (paper's `kt`) | 0.0078 1/h (Wistar); 0.0193 1/h (Iar:COP) | Table 2, row `k t` |
| `logitfrel`, `logitfrel2` (paper's `NR`, `DR`, `ER`) | 0.18 / 0.28 / 0.54 (Wistar); 0.08 / 0.43 / 0.49 (Iar:COP) | Table 2, rows `N R`, `D R`, `E R` |
| `ltlag`, `ltlag2` | 0.47 / 3.61 h (Wistar); 0.35 / 2.58 h (Iar:COP) | Table 2, rows `t lag,d`, `t lag ,e` |
| erosion chain length = 2 | 2 (fixed) | Table 2, row `ES n`; Results 2.4 |
| `lkin_tt`, `lkout_tt` | 0.68 / 0.16 (Wistar); 0.35 / 0.06 (Iar:COP) | Table 4, rows `k in`, `k out` |
| `lhill` | 2.00 / 2.02 / 2.00 / 3.18 | Table 4, row `h` |
| `lemax` | 303.77 / 183.50 / 380.00 / 634.50 | Table 4, row `E max` |
| `lec50` | 3.48 / 6.17 / 1.80 / 3.34 ng/mL | Table 4, row `EC 50` |
| `lkin_moderator1` (paper's `kf,on`) | 0.29 / 0.14 / 0.40 / 0.083 | Table 4, row `k f,on` |
| `lkout_moderator1` (paper's `kf,off`) | 0.059 / 0.02 / 0.04 / 0.45 1/h | Table 4, row `k f,off` |
| `propSd`, `propSd_TT` | `fixed(0)` | not reported anywhere in the source |

Where four values are listed they are in the order Group 1 / Group 3 / Group 4 /
Group 6, matching the column order of Table 4.

# Virtual cohort

Neither between-animal variability nor a residual error model is reported: Lee
2018 fits each group's mean profile deterministically in Berkeley Madonna by the
Runge-Kutta method (Methods 4.7), and every `+/-` in its tables is the standard
error of the estimate across n = 5 animals. There is therefore nothing to
simulate stochastically - the models carry no `eta` terms - and each group is a
single deterministic typical-value trajectory.

```{r cohort}
# Fine early so the ~0.18 h absorption peak and the ~3 h testosterone flare are
# resolved; coarse over the 14-day tail.
grid <- sort(unique(c(seq(0, 2, by = 0.005), seq(2, 24, by = 0.05), seq(24, 336, by = 0.5))))

# Dose compartments per group. The depot model splits one administered amount
# across three release sections (Equations 4-6), so it carries three
# simultaneous dose records and the f() fractions in the model do the splitting.
spec <- tibble::tibble(
  model = model_names,
  group = factor(
    c("Group 1: IV solution", "Group 3: SC solution", "Group 4: SR depot (Wistar)", "Group 6: SR depot (Iar:COP)"),
    levels = c("Group 1: IV solution", "Group 3: SC solution", "Group 4: SR depot (Wistar)", "Group 6: SR depot (Iar:COP)")
  ),
  amt = c(dose_iv, dose_sc, dose_sc, dose_sc)
)
dose_cmts <- list(
  "central",
  "depot",
  c("depot", "depot2", "transit1"),
  c("depot", "depot2", "transit1")
)

make_events <- function(cmts, amt) {
  ev <- rxode2::et(amt = amt, cmt = cmts[1])
  for (cc in cmts[-1]) ev <- rxode2::et(ev, amt = amt, cmt = cc)
  # `TT` is both an ODE state and a declared endpoint, so it is a valid
  # observation compartment for this two-endpoint model without injecting a new
  # slot the way an algebraic observable such as `Cc` would. rxode2 returns
  # every model variable as a column at these rows, `Cc` included.
  rxode2::et(ev, grid, cmt = "TT")
}
events <- Map(make_events, dose_cmts, spec$amt)
names(events) <- model_names
```

# Simulation

```{r simulate}
solve_one <- function(nm) {
  as.data.frame(rxode2::rxSolve(
    uis[[nm]], events[[nm]],
    atol = 1e-12, rtol = 1e-10,
    # rxode2's ODE -> linCmt() auto-conversion mis-maps dvid for multi-output
    # models; see known-vignette-failure-patterns pattern 5b.
    useLinCmt = FALSE
  )) |>
    dplyr::mutate(model = nm)
}
# Kept both per-model (the four models declare different state sets, so binding
# them introduces NA columns) and stacked for plotting.
sims <- lapply(model_names, solve_one)
names(sims) <- model_names
sim <- dplyr::bind_rows(sims) |>
  dplyr::left_join(spec, by = "model") |>
  dplyr::mutate(day = time / 24)
```

# Replicate published figures

## Figure 2, leuprolide

Figure 2 plots leuprolide and testosterone for Groups 1, 3, 4 and 6 (panels A-D).

```{r figure-2-pk}
# Replicates the left-hand leuprolide panels of Figure 2 of Lee 2018.
sim |>
  dplyr::filter(Cc > 1e-4) |>
  ggplot(aes(time, Cc)) +
  geom_line() +
  facet_wrap(~group, scales = "free_x") +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Plasma leuprolide (ng/mL)",
    title = "Figure 2 (left panels) - leuprolide",
    caption = "Replicates the leuprolide panels of Figure 2 of Lee 2018."
  )
```

Note the x-axis unit. Figure 2's leuprolide panels are labelled "Time (day)",
but for the two solution groups they are in **hours**: the drawn line falls five
log10 units across eight x-units, i.e. a slope of 1.44 per unit, which is the
reported elimination rate constant `k` = 1.42 1/h and not any per-day quantity;
and panel B's peak sits at x ~ 0.18, which is the analytic `Tmax` computed
below. This is recorded as an erratum.

## Figure 2, testosterone

```{r figure-2-pd}
# Replicates the right-hand testosterone panels of Figure 2 of Lee 2018.
sim |>
  ggplot(aes(day, TT)) +
  geom_line() +
  geom_hline(aes(yintercept = rbase_tt), linetype = "dashed", colour = "grey50") +
  facet_wrap(~group) +
  scale_y_log10() +
  labs(
    x = "Time (day)", y = "Plasma testosterone (ng/mL)",
    title = "Figure 2 (right panels) - testosterone",
    caption = paste(
      "Replicates the testosterone panels of Figure 2 of Lee 2018.",
      "Dashed line is the published baseline R0 (Equations 12 and 13)."
    )
  )
```

The shape is the paper's: a sharp flare-up within the first hours, a nadir below
baseline around day 1-2, and recovery to a plateau. The **amplitude of the
flare** is not reproduced; see Errata item 2.

## Figure 5, model schematic

Figure 5's schematic is reproduced structurally by the compartment metadata:

```{r schematic}
purrr_free <- function(u) {
  cd <- u$compartmentData
  tibble::tibble(State = names(cd), Analyte = vapply(cd, `[[`, character(1), "analyte"))
}
purrr_free(uis[["Lee_2018_leuprolide_sr_rat"]]) |>
  knitr::kable(caption = "States of the sustained-release model (Figure 5b of Lee 2018).")
```

# Validation

## Check 1 - release-section fractions and Equation (7)

The three release fractions are held on a 2-simplex by a stick-breaking logit
pair rather than as three free parameters, so the first thing to prove is that
the encoding returns Table 2 unchanged and that Equation (7) holds identically.

```{r check-fractions}
fracs <- sim |>
  dplyr::filter(model %in% c("Lee_2018_leuprolide_sr_rat", "Lee_2018_leuprolide_sr_cancer_rat")) |>
  dplyr::group_by(group) |>
  dplyr::summarise(NR = dplyr::first(frel), DR = dplyr::first(frel2), ER = dplyr::first(frel3), .groups = "drop") |>
  dplyr::mutate(sum = NR + DR + ER)
knitr::kable(fracs, digits = 10, caption = "Release fractions recovered from the stick-breaking encoding.")

published_fracs <- tibble::tibble(NR = c(0.18, 0.08), DR = c(0.28, 0.43), ER = c(0.54, 0.49))
stopifnot(
  # Deterministic algebra, so an exact comparison is the right assertion.
  max(abs(fracs$NR - published_fracs$NR)) < 1e-12,
  max(abs(fracs$DR - published_fracs$DR)) < 1e-12,
  max(abs(fracs$ER - published_fracs$ER)) < 1e-12,
  max(abs(fracs$sum - 1)) < 1e-12
)
```

## Check 2 - mass balance of the three-section depot

All of `NR`, `DR` and `ER` eventually reach the central compartment and nothing
is lost on the way, so `AUC(0-inf) * CL` must equal the administered amount
exactly. This is a real test of the depot plumbing: a mis-wired transit chain, a
dropped `f()` fraction or a lag applied to the wrong state all break it. Because
the depot's erosion chain has a mean transit time of `2 / kt` = `r round(2 / 0.0078 / 24, 1)` days
for the Wistar group, a large part of the dose is still unreleased at day 14, so
the remaining amount is added back explicitly.

```{r check-mass-balance}
state_cols <- c("depot", "depot2", "transit1", "transit2", "central")
mass <- lapply(model_names, function(nm) {
  r <- sims[[nm]]
  last_row <- r[nrow(r), , drop = FALSE]
  tibble::tibble(
    model = nm,
    auc_336 = sum(diff(r$time) * (head(r$Cc, -1) + tail(r$Cc, -1)) / 2),
    cl = r$cl[1],
    # Drug still in the depot / transit / central states at day 14 will all
    # eventually be cleared, so it contributes `amount / CL` to AUC(0-inf).
    remaining = sum(unlist(last_row[intersect(state_cols, names(last_row))]))
  )
}) |>
  dplyr::bind_rows() |>
  dplyr::left_join(spec, by = "model") |>
  dplyr::mutate(
    auc_inf = auc_336 + remaining / cl,
    recovered = auc_inf * cl,
    ratio = recovered / amt
  )

mass |>
  dplyr::select(group, `AUC 0-336 h` = auc_336, `AUC 0-inf` = auc_inf, `Recovered (ng)` = recovered, `Dose (ng)` = amt, Ratio = ratio) |>
  knitr::kable(digits = c(0, 3, 3, 1, 1, 6), caption = "Mass balance: AUC(0-inf) x CL against the administered amount.")

stopifnot(max(abs(mass$ratio - 1)) < 1e-3)
```

## Check 3 - analytic PK identities against Table 1

`t1/2` is `log(2) * Vd / CL` and, for the subcutaneous solution,
`Tmax = log(ka / kel) / (ka - kel)`. Neither depends on the dose, so unlike the
AUC comparison below these are independent of the back-solved amount.

```{r check-analytic}
kel_iv <- 248.61 / 192.95
kel_sc <- 514.46 / 487.40
ka_sc <- 16.67

analytic <- tibble::tibble(
  Quantity = c("t1/2, IV solution (h)", "t1/2, SC solution (h)", "Tmax, SC solution (h)"),
  Model = c(log(2) / kel_iv, log(2) / kel_sc, log(ka_sc / kel_sc) / (ka_sc - kel_sc)),
  Published = c(0.52, 0.66, NA_real_),
  Source = c("Table 1, IV, row t 1/2", "Table 1, SC, row t 1/2", "not reported; Figure 2B peak sits at x ~ 0.18")
) |>
  dplyr::mutate(`% diff` = (Model - Published) / Published * 100)
knitr::kable(analytic, digits = 4, caption = "Analytic PK identities against Table 1.")

# Deterministic; 6% admits the internal inconsistency of Table 1 (its k, CL/Vd
# and t1/2 columns disagree by up to 10% because each is a mean of per-animal
# values), and still goes red on any transcription error in CL or Vd.
stopifnot(max(abs(analytic$`% diff`), na.rm = TRUE) < 6)
```

Table 1 reports `k`, `CL` and `Vd` as separate per-animal means that are not
mutually consistent (`CL / Vd` = `r round(kel_iv, 3)` 1/h for the IV group
against a reported `k` of 1.42 1/h and a `t1/2` of 0.52 h, i.e. 1.33 1/h).
Equation (2) writes the elimination term as `CL / Vd`, so that is what the model
uses and `k` is left derived.

## Check 4 - observed-Cmax comparison

Table 1's Cmax values are the highest **observed** samples, and the earliest
sample was at 0.25 h. The like-for-like model quantity is therefore the
concentration at 0.25 h, not the peak of the continuous curve (which for an IV
bolus is the back-extrapolated `C0` = `dose / Vd`).

```{r check-cmax}
c025 <- sim |>
  dplyr::filter(model %in% model_names[1:2], abs(time - 0.25) < 1e-9) |>
  dplyr::select(group, Cc)
cmax_tab <- c025 |>
  dplyr::mutate(
    `Published Cmax` = c(97.10, 41.53),
    `% diff` = (Cc - `Published Cmax`) / `Published Cmax` * 100
  ) |>
  dplyr::rename("Model C at 0.25 h" = Cc)
knitr::kable(cmax_tab, digits = 3, caption = "Model concentration at the first sampling time against Table 1 Cmax.")

stopifnot(max(abs(cmax_tab$`% diff`)) < 12)
```

## Check 5 - PKNCA against Table 1

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(model %in% model_names[1:2], time <= 12) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(id = as.integer(factor(model))) |>
  dplyr::select(id, time, Cc, group)

# The time-zero row is already present from the solve (Cc = C0 for the IV bolus,
# 0 for the subcutaneous route), so the defensive row below is a no-op that
# distinct() discards; it is kept so the guarantee is explicit.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, group) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, group, time, .keep_all = TRUE) |>
  dplyr::arrange(id, group, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | group + id)

dose_df <- sim_nca |>
  dplyr::distinct(id, group) |>
  dplyr::mutate(time = 0, amt = ifelse(id == 1L, dose_iv, dose_sc))
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | group + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r nca-compare}
published <- tibble::tibble(
  group = levels(spec$group)[1:2],
  # Table 1 Cmax for the IV group is the highest observed sample (0.25 h), not
  # the back-extrapolated C0 that PKNCA reports here, so it is left out of the
  # comparison; Check 4 makes the like-for-like comparison instead.
  cmax = c(NA_real_, 41.53),
  auclast = c(103.44, 51.20),
  aucinf.obs = c(105.50, 53.33),
  half.life = c(0.52, 0.66)
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "group",
  # half.life = TRUE also emits span.ratio / r.squared / lambda.z diagnostics;
  # restrict to the parameters the paper actually reports.
  params = c("cmax", "auclast", "aucinf.obs", "half.life"),
  units = c(cmax = "ng/mL", auclast = "h*ng/mL", aucinf.obs = "h*ng/mL", half.life = "h"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Simulated vs Table 1 NCA. * differs from reference by >20%.")
attr(cmp, "footnote")
```

**The two AUC rows are a transcription check, not an independent gate.** The
administered amount was back-solved as `AUC0-inf x CL` from this very table, so
`AUC0-inf = dose / CL` holds by construction; what the row actually proves is
that the model's `CL` was transcribed correctly and that the solver conserves
mass. The `half.life` and subcutaneous `cmax` rows are independent of the dose
and are the informative ones.

## Check 6 - closed-form steady state of the feedback loop

Away from drug (`ce` -> 0), Equations (14) and (15) have the joint fixed point

$$ R_{ss} = \sqrt{\frac{k_{in} \, k_{f,on}}{k_{out} \, k_{f,off}}}, \qquad F_{ss} = \frac{k_{f,on}}{k_{f,off} R_{ss}} $$

obtained by setting both derivatives to zero. Comparing that closed form against
the numerically integrated trajectory tests that the two coupled ODEs were
encoded as written - a sign error, a swapped rate constant or a missing division
by `R` all move the fixed point.

```{r check-ss}
ss <- sim |>
  dplyr::group_by(group) |>
  dplyr::summarise(
    closed_form = sqrt(dplyr::first(kin_tt) * dplyr::first(kin_moderator1) /
      (dplyr::first(kout_tt) * dplyr::first(kout_moderator1))),
    solved_14d = dplyr::last(TT),
    .groups = "drop"
  ) |>
  dplyr::mutate(ratio = solved_14d / closed_form)
knitr::kable(ss, digits = 4, caption = "Closed-form joint steady state against the trajectory at day 14.")

# Groups 1, 3 and 6 have essentially no drug left at day 14 and must sit on the
# fixed point exactly. Group 4's depot is still releasing (erosion mean transit
# time ~10.7 days), so its trajectory is legitimately still above the drug-free
# fixed point and is excluded.
stopifnot(max(abs(ss$ratio[c(1, 2, 4)] - 1)) < 1e-4)
```

The same closed form is what localises the Group 6 problem: it reproduces the
Figure 2 plateaus of all three Wistar groups but not Group 6's. See Errata
item 3.

## Check 7 - drug-free behaviour against the vehicle groups

Groups 2 and 5 received vehicle only and showed no change in testosterone over
the whole experiment ("The dispersion solution did not induce any significant
changes for the entire experimental period", Results 2.3). Solving each model
with no dose is the corresponding test.

```{r check-vehicle}
vehicle <- lapply(model_names, function(nm) {
  as.data.frame(rxode2::rxSolve(uis[[nm]], rxode2::et(grid, cmt = "TT"), atol = 1e-12, rtol = 1e-10, useLinCmt = FALSE)) |>
    dplyr::mutate(model = nm)
}) |>
  dplyr::bind_rows() |>
  dplyr::left_join(spec, by = "model") |>
  dplyr::group_by(group) |>
  dplyr::summarise(
    `R0 (Eq 12/13)` = dplyr::first(rbase_tt),
    `kin / kout` = dplyr::first(kin_tt) / dplyr::first(kout_tt),
    `TT at day 14` = dplyr::last(TT),
    `% drift from R0` = (dplyr::last(TT) - dplyr::first(rbase_tt)) / dplyr::first(rbase_tt) * 100,
    .groups = "drop"
  )
knitr::kable(vehicle, digits = 3, caption = "Drug-free trajectory: published baseline against where the coupled system settles.")
```

The drug-free baseline is **not** flat for any group, because the published
`R0`, `kin / kout` and `sqrt(kin kf,on / (kout kf,off))` are three different
numbers in every column of Table 4. For the three Wistar groups the drift is
5-50% over 14 days and the settled level still brackets the observed post-nadir
plateau; for Group 6 the system falls fourfold. This is a property of the
published parameter set, not of the encoding - see Errata items 3 and 4.

## Documented deviations

The one aspect of the paper that this implementation does **not** reproduce is
the amplitude of the testosterone flare. It is recorded here rather than tuned
away, and is deliberately excluded from the assertions above.

```{r deviations}
flare <- sim |>
  dplyr::group_by(group) |>
  dplyr::summarise(
    `Model peak TT (ng/mL)` = max(TT),
    `at (h)` = time[which.max(TT)],
    .groups = "drop"
  ) |>
  dplyr::mutate(
    `Figure 2 fitted peak (read)` = c(65, 45, 110, 45),
    `Ratio` = `Model peak TT (ng/mL)` / `Figure 2 fitted peak (read)`
  )
knitr::kable(flare, digits = 2, caption = "KNOWN DEVIATION: flare amplitude. Not asserted; see Errata item 2.")
```

# Assumptions and deviations (Errata)

1. **Administered amount is back-solved, not reported.** Body weight appears
   nowhere in the paper, so the 0.1 mg/kg dose could not be converted to an
   amount from the text. It was derived from the paper's own
   `CL = Dose / AUC0-inf` definition, independently per route, giving 26,228 ng
   (IV) and 27,436 ng (SC) and implied body weights of 262 g and 274 g. The
   agreement between the two routes is the evidence that the derivation is
   sound. A consequence: the `AUC` rows of the PKNCA comparison are circular by
   construction and are labelled as such.

2. **The flare-up amplitude is over-predicted by a factor of 2-6 for the three
   Wistar groups.** Equations (14) and (16) with the Table 4 values give peak
   testosterone of 396 / 198 / 412 / 101 ng/mL for Groups 1 / 3 / 4 / 6, and the
   maximum prediction at an actual sampling time is 412 ng/mL. Figure 3D, the
   paper's own weighted-residual plot, has predicted testosterone on its x-axis
   and it tops out at about 95 ng/mL; the fitted curves in Figure 2 peak at
   roughly 65 / 45 / 110 / 45 ng/mL. The one free quantity is the unstated
   initial condition of the feedback pool, `F(0)`, which the paper never gives.
   It cannot explain the pattern: the peak is very nearly proportional to `F(0)`
   while the nadir and the plateau are completely independent of it, so matching
   each group in turn would need `F(0)` = 0.16 / 0.23 / 0.27 / 0.44, not one
   value. `F(0) = 1` is used because it is the value at which Equation (14)
   collapses exactly onto the paper's own baseline model, Equation (11), which
   is how the paper describes the drug-effect model being constructed. Nothing
   was tuned. Everything downstream of the flare - nadir depth, nadir timing and
   recovered plateau - does reproduce for the Wistar groups.

3. **The Group 6 (Iar:COP) feedback constants do not reproduce the paper's own
   Figure 2D.** The joint fixed point `sqrt(kin kf,on / (kout kf,off))` gives
   4.57 / 5.45 / 6.52 ng/mL for the three Wistar groups, matching the plateaus
   drawn in Figure 2A-C (about 4.5-5, 5.5 and 7). For Group 6 the same closed
   form gives 1.04 ng/mL against a plateau of about 4.2 in Figure 2D and an
   observed 3-14 day mean of 4.39 ng/mL reported in Results 2.3. Group 6 is also
   the only column of Table 4 whose `kf,on` is smaller than its `kf,off` (by a
   factor of 5.4, where the other three have `kf,on` larger by 5-10x) and the
   only one whose `kf,on` standard error (0.14) exceeds its point estimate
   (0.083). The published values are transcribed unchanged and the
   inconsistency is flagged in that model file's `description`.

4. **`R0` is not the steady state of the model it initialises.** Equation (11)
   is at rest when `R = kin / kout`, which is 4.25 for Wistar against the
   `R0 = 4.353` of Equation (12) - consistent within 2.4% - but 5.83 for
   Iar:COP against the `R0 = 4.094` of Equation (13), a 42% disagreement. The
   full coupled system settles at a third value again. All three published
   numbers are kept as printed; the drug-free drift this produces is quantified
   in Check 7.

5. **`F(0) = 1`** is an assumption, as above. The alternative of starting the
   pool at its own quasi-steady state `kf,on / (kf,off R0)` shifts the flare
   peak by about 13% for Group 1 and leaves the nadir and plateau unchanged.

6. **No between-animal variability and no residual error model are reported.**
   Every `+/-` in Tables 1-4 is the standard error of a parameter estimate
   across n = 5 animals, not a variance component: Methods 4.7 fits each group's
   mean profile deterministically in Berkeley Madonna. `propSd` and `propSd_TT`
   are therefore `fixed(0)` rather than invented, and the models carry no `eta`
   terms.

7. **Figure 2's leuprolide x-axis is mislabelled for the solution groups.** It
   reads "Time (day)" but is in hours - the fitted line's slope is 1.44 per
   x-unit, which is the reported `k` = 1.42 1/h, and panel B's peak at
   x ~ 0.18 is the analytic `Tmax` of 0.177 h. The depot panels (C and D), which
   run to 14 days, are genuinely in days.

8. **Table 3's AUEC values are not reproduced and no attempt is made to.** They
   are computed from observed data by subtracting a baseline AUEC and splitting
   the remainder into positive and negative segments (Discussion, Methods 4.6),
   but the segmentation convention and the units are not specified. The paper's
   own fitted curves recover *above* baseline after the nadir, so the model's
   negative area is roughly fifty times smaller than the published figures; a
   numeric comparison would measure the undefined convention, not the model.

9. **`kf,on` units.** Table 4 heads both the `kf,on` and `kf,off` rows "h - 1".
   That is right for `kf,off`, but Equation (15) divides `kf,on` by a
   concentration, so `kf,on` must carry ng/mL/h for `dF/dt` to be a pure 1/h
   rate. The `label()` states the units the equation requires and records what
   the table prints.

10. **Disposition is shared across groups, as the paper specifies.** The two
    depot models do not re-estimate `ka`, `CL/F` or `Vd/F`; Methods 4.7 builds
    the depot model "based on the PK parameters such as ka, CL, and Vd in
    solution-administered groups", so those three are carried from Group 3
    unchanged, including into the Iar:COP model. The paper notes that PK did not
    differ significantly between strains while `Emax` and `EC50` did.

11. **Naming.** The paper's `kd` and `kt` are encoded as the canonical `lka2`
    and `lktr`: Results 2.4 states that both "are rate constants that include
    both the release process in the biological systems and the absorption
    process of the released drug", i.e. apparent first-order input rate
    constants, which is exactly what those canonicals denote. The three release
    fractions `NR` / `DR` / `ER` are held by a stick-breaking `logitfrel` /
    `logitfrel2` pair so that Equation (7) holds identically; Check 1 proves the
    printed values are recovered. The feedback pool is the canonical
    `moderator1` compartment with `kin_moderator1` / `kout_moderator1` rate
    constants, keeping it clear of the testosterone pool's `kin_tt` / `kout_tt`.
