---
title: "Roscovitine (Leven 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Roscovitine (Leven 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Leven_2020_roscovitine"))$reference`
* Description: `r rxode2::rxode(readModelDb("Leven_2020_roscovitine"))$description`
* Article (open access): <https://doi.org/10.3390/pharmaceutics12111087>

Roscovitine (seliciclib) is an oral purine-analogue protein-kinase inhibitor
investigated in cystic fibrosis (CF) for its effects on F508del-CFTR
degradation and on alveolar-macrophage bactericidal activity. Leven 2020
characterised its pharmacokinetics and that of its main carboxylate metabolite
M3 after the first oral dose in the ROSCO-CF phase IIa trial. The final model
has three unusual features, all reproduced here:

1. **A dose-dependent, saturable first-pass effect.** The fraction of the dose
   reaching the systemic circulation as roscovitine is
   `fdose = Dose / (Dose + D50)` (Equation 7); the remainder, `1 - fdose`,
   enters the circulation as pre-formed M3.
2. **A sum-of-inverse-Gaussian absorption input** without any first-order
   absorption step (Equations 4-6): two parallel roscovitine inputs weighted
   `pi1` and `1 - pi1`, plus one M3 input whose shape is fixed to the first
   roscovitine input (`TM3max = T1max`, `CVM3 = CV1`, Section 3.2).
3. **No roscovitine elimination other than systemic conversion to M3**
   (`kmet`), and one apparent volume `V/F` shared by both molecules, because
   neither a parent elimination route nor a separate M3 volume was identifiable
   from oral data (Section 3.2).

The paper's Appendix A prints the complete MLXTRAN structural model, which is
what this implementation follows line by line.

# Population

Twenty-three adults with CF (two CF-causing mutations including at least one
F508del, FEV1 >= 40%, chronic *Pseudomonas aeruginosa* infection) received a
single morning oral dose of 200 mg (n = 9), 400 mg (n = 7) or 800 mg (n = 7)
roscovitine as 200 mg hard gelatin capsules, with PK sampled pre-dose and at
0.5, 1, 2, 4, 6, 8 and 12 h after the first dose (Sections 2.2-2.3). Median
(range) age was 33 (22-51) years, weight 58 (48-90) kg, height 166 (158-182)
cm and MDRD GFR 113 (61-202) mL/min; 10 were female and 15 took a proton-pump
inhibitor (PPI; Table 2). The data comprised 138 roscovitine (19 BLQ) and 138
M3 (9 BLQ) concentrations (Table 1).

The same information is available programmatically via
`readModelDb("Leven_2020_roscovitine")()$population`.

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `led50` (D50) | log(1190 umol) | Table 3 |
| `logitfrel` (pi1) | logit(0.285) | Table 3; logit-normal, Section 2.5 |
| `ltpeak_ig1` (T1max) | log(0.676 h) | Table 3 |
| `e_conmed_ppi_tpeak_ig1` | 0.680 | Table 3, 'beta (PPI = 1) on T1max'; Equation 9 |
| `ldtpeak_ig2` (dT2max) | log(1.04 h) | Table 3; T2max = T1max + dT2max, Section 3.2 |
| `lcv_ig1`, `lcv_ig2` (CV1, CV2) | log(0.542), log(0.354) | Table 3 |
| `lvc` (V/F) | log(62.2 L) | Table 3 |
| `e_ht_vc` | 6.47 | Table 3; Equation 10 with reference height 170 cm (Section 3.2) |
| `lk12`, `lk21` | log(1.82), log(0.768) 1/h | Table 3 |
| `lkmet` | log(2.07 1/h) | Table 3 |
| `lkel_m3` (ke) | log(2.58 1/h) | Table 3 |
| IIV variances | omega^2 of the Table 3 SDs | Table 3 (D50 0.689, T1max 0.452, dT2max 0.647, CV1 0.426, CV2 0.547, V/F 0.678, ke 0.393) |
| IIV covariances | corr x omega x omega | Table 3 (T1max-CV2 -0.839; dT2max-CV1 0.624) |
| `propSd` | 0.297 | Table 3, 'Roscovitine b1' |
| `addSd_m3`, `propSd_m3` | 9.20 nmol/L, 0.271 | Table 3, 'M3 a2', 'M3 b2'; combined error `sqrt(a^2 + b^2 f^2)`, Section 2.5 |
| `fdose = Dose / (Dose + D50)` | n/a | Equation 7; Appendix A `Fr` |
| `MAT = Tmax / (sqrt(1 + 9/4 CV^4) - 3/2 CV^2)` | n/a | Equation 6; Appendix A `MAT_1`..`MAT_3` |
| inverse-Gaussian density `IG(t)` | n/a | Equation 5; Appendix A `inv_gauss_1`..`inv_gauss_3` |
| `d/dt(central)`, `d/dt(peripheral1)`, `d/dt(central_m3)` | n/a | Figure 2; Appendix A `dArosco`, `dAroscop`, `dAM3` |
| `Cc = central / vc * 1000`, `Cc_m3 = central_m3 / vc * 1000` | n/a | Appendix A `C_rosco`, `C_M3` (nmol/L) |

## Implementation notes

* **Dose units.** The model is dosed in **micromoles** of roscovitine and
  predicts both analytes in **nmol/L**, exactly as the paper fitted it
  (Section 2.3: doses converted to umol, concentrations to nmol/L, molecular
  weights 354.45 g/mol for roscovitine and 368.43 g/mol for M3). D50 is
  therefore in umol: a 200 mg dose is `200 / 354.45 * 1000 = 564.3` umol.
* **The input function.** Monolix evaluates the inverse-Gaussian inputs from
  the dose amount (`amtDose`) and the time since dosing. rxode2 exposes the
  same two quantities as `podo(depot)` and `tad(depot)`. The dose is placed in
  `depot`, which is emptied at exactly the prescribed input rate, so the total
  amount in the system is conserved; the mass-balance check below confirms
  that each dose is delivered once and completely.
* **Repeated dosing.** The model was fitted to the first dose only. With
  `podo()` / `tad()` a new dose restarts the input clock, which is accurate as
  long as the previous dose has been absorbed (the slower density has a mean
  absorption time of about 2 h at typical values, so this holds for the
  once-daily ROSCO-CF regimen).

# Molecular weights and dose conversion

```{r constants}
MW_ROSCO <- 354.45 # g/mol, Section 2.3
MW_M3 <- 368.43 # g/mol, Section 2.3
mg_to_umol <- function(mg) mg / MW_ROSCO * 1000

mod <- rxode2::rxode2(readModelDb("Leven_2020_roscovitine"))
# The explicit ODEs must be kept: a model with a volume and micro-constants
# must not be silently converted to a closed-form linear compartment model.
stopifnot(is.null(mod$linCmt))
ini_vals <- mod$theta
```

# Event-table builder

```{r event_builder}
# Event tables are data frames, so the covariate columns survive. The model
# declares two endpoints, so observation rows sit on a real ODE state
# (cmt = "central") and nominate an endpoint with dvid = 1; both observables,
# Cc and Cc_m3, come back as columns on every observation row.
build_events <- function(cohort, obs_times) {
  dose <- cohort |>
    dplyr::transmute(id, time = 0, amt = mg_to_umol(dose_mg), evid = 1L,
                     cmt = "depot", dvid = NA_integer_, HT, CONMED_PPI)
  obs <- cohort |>
    dplyr::select(id, HT, CONMED_PPI) |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(amt = 0, evid = 0L, cmt = "central", dvid = 1L) |>
    dplyr::select(id, time, amt, evid, cmt, dvid, HT, CONMED_PPI)
  dplyr::bind_rows(dose, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}
```

# Deterministic checks

## First-pass fraction

The Results (Section 3.2) state that D50 = 1190 umol gives
`Fr` = 0.32, 0.49 and 0.65 at 200, 400 and 800 mg. The model's own `fdose`
at the dose time is compared against those printed values; the tolerance is
half of the last printed digit.

```{r fdose_check}
typ <- rxode2::zeroRe(mod)
doses_mg <- c(200, 400, 800)
typ_cohort <- tibble::tibble(id = 1:3, dose_mg = doses_mg, HT = 170, CONMED_PPI = 0)
typ_times <- c(0, exp(seq(log(0.001), log(0.05), length.out = 40)), seq(0.1, 96, by = 0.05))
typ_sim <- rxode2::rxSolve(typ, build_events(typ_cohort, typ_times),
                           returnType = "data.frame", atol = 1e-12, rtol = 1e-10,
                           maxsteps = 1e6) |>
  dplyr::left_join(typ_cohort, by = "id")

fdose_tab <- typ_sim |>
  dplyr::group_by(dose_mg) |>
  dplyr::summarise(fdose = dplyr::last(fdose), .groups = "drop") |>
  dplyr::mutate(published = c(0.32, 0.49, 0.65))
knitr::kable(fdose_tab |> dplyr::rename("Dose (mg)" = dose_mg, "Model fdose" = fdose,
                                        "Published Fr" = published),
             digits = 4, caption = "Fraction of the dose escaping first-pass conversion.")
stopifnot(all(abs(fdose_tab$fdose - fdose_tab$published) < 0.005))
```

## Mass balance

Roscovitine leaves the body only by conversion to M3, and M3 only by `ke`.
Therefore, for any subject, the whole dose must be eliminated as M3 and the
fraction `fdose` must pass through systemic roscovitine:

* `ke * V * AUC0-inf(M3) / 1000 = Dose` (umol),
* `kmet * V * AUC0-inf(roscovitine) / 1000 = fdose * Dose`.

Both sides come from the same solve, so the only discrepancy allowed is
numerical integration and trapezoidal error. An input function that delivered
nothing, delivered a dose twice, or put the pre-systemic fraction in the wrong
compartment would fail these by tens of percent.

```{r mass_balance}
trap <- function(x, y) sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)
mb <- typ_sim |>
  dplyr::group_by(dose_mg) |>
  dplyr::summarise(
    dose_umol = mg_to_umol(dplyr::first(dose_mg)),
    fdose = dplyr::last(fdose),
    auc_m3 = trap(time, Cc_m3),
    auc_rosco = trap(time, Cc),
    depot_end = dplyr::last(depot),
    vc = dplyr::last(vc), kel_m3 = dplyr::last(kel_m3), kmet = dplyr::last(kmet),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    m3_recovered = kel_m3 * vc * auc_m3 / 1000 / dose_umol,
    rosco_recovered = kmet * vc * auc_rosco / 1000 / (fdose * dose_umol)
  )
knitr::kable(mb |> dplyr::select(dose_mg, m3_recovered, rosco_recovered, depot_end) |>
               dplyr::rename("Dose (mg)" = dose_mg,
                             "M3 eliminated / dose" = m3_recovered,
                             "Systemic roscovitine / (fdose x dose)" = rosco_recovered,
                             "Depot at 96 h (umol)" = depot_end),
             digits = 6, caption = "Typical-value mass balance (reference height 170 cm, no PPI).")
stopifnot(
  all(abs(mb$m3_recovered - 1) < 1e-3),
  all(abs(mb$rosco_recovered - 1) < 1e-3),
  all(abs(mb$depot_end) < 1e-6 * mb$dose_umol)
)
```

# Virtual cohort

The paper's Monte Carlo simulations used 1000 patients per scenario but do not
state the covariate values. Simulating at the **reference height of 170 cm**,
with PPI use drawn at the observed prevalence (15 of 23), is consistent with
the published medians (tables below), whereas a cohort whose heights span the
observed 158-182 cm range is expected to give higher roscovitine and M3
exposure because V/F scales as `(HT / 170)^6.47` and the observed median height
(166 cm) lies below the reference (see the sensitivity section at the end).
The primary cohort therefore uses 170 cm; each dose arm has 200 virtual
patients.

```{r cohort}
rxode2::rxSetSeed(2020)
set.seed(2020)
n_per_arm <- 200

# Heights for the sensitivity cohort: normal around the observed median, redrawn
# (never clamped) until inside the observed 158-182 cm range (Table 2).
draw_height <- function(n) {
  out <- numeric(0)
  while (length(out) < n) {
    x <- rnorm(n, 166, 7)
    out <- c(out, x[x >= 158 & x <= 182])
  }
  out[seq_len(n)]
}

make_cohort <- function(n, dose_mg, ppi = NULL, height = NULL, id_offset = 0L) {
  tibble::tibble(
    id = id_offset + seq_len(n),
    dose_mg = dose_mg,
    HT = if (is.null(height)) 170 else height(n),
    CONMED_PPI = if (is.null(ppi)) rbinom(n, 1, 15 / 23) else ppi
  )
}

dose_cohort <- dplyr::bind_rows(
  make_cohort(n_per_arm, 200, id_offset = 0L),
  make_cohort(n_per_arm, 400, id_offset = 1000L),
  make_cohort(n_per_arm, 800, id_offset = 2000L)
) |>
  dplyr::mutate(treatment = paste(dose_mg, "mg"))

obs_times <- sort(unique(c(seq(0, 12, by = 0.1))))
dose_sim <- rxode2::rxSolve(mod, build_events(dose_cohort, obs_times),
                            returnType = "data.frame") |>
  dplyr::left_join(dose_cohort |> dplyr::select(id, dose_mg, treatment), by = "id")
```

# Simulated profiles by dose (replicates Figure 7)

```{r figure7, fig.height = 6}
band <- dose_sim |>
  dplyr::select(treatment, time, Roscovitine = Cc, `M3 metabolite` = Cc_m3) |>
  tidyr::pivot_longer(c(Roscovitine, `M3 metabolite`), names_to = "analyte", values_to = "conc") |>
  dplyr::group_by(treatment, analyte, time) |>
  dplyr::summarise(p10 = quantile(conc, 0.1), p50 = median(conc), p90 = quantile(conc, 0.9),
                   .groups = "drop") |>
  dplyr::mutate(analyte = factor(analyte, c("Roscovitine", "M3 metabolite")),
                treatment = factor(treatment, c("200 mg", "400 mg", "800 mg")))

ggplot(band, aes(time, p50)) +
  geom_ribbon(aes(ymin = p10, ymax = p90), fill = "steelblue", alpha = 0.3) +
  geom_line(colour = "steelblue4") +
  facet_grid(analyte ~ treatment, scales = "free_y") +
  labs(x = "Time (h)", y = "Concentration (nmol/L)",
       caption = "Median and 10th-90th percentiles; replicates Figure 7 of Leven 2020.")
```

As in Figure 7, roscovitine exposure rises more than proportionally with dose
while the M3 peak, which precedes the roscovitine peak because its input shares
the faster first density, rises less than proportionally. Over an infinite
horizon M3 exposure is exactly dose-proportional (every umol of roscovitine is
eventually eliminated as M3, as the mass-balance check shows); the
less-than-proportional AUC0-12h arises because at higher doses a larger share
of the dose passes through systemic roscovitine and its peripheral compartment
before becoming M3, so more of the M3 is formed after 12 h. Both statements are
asserted below the PKNCA table.

# PKNCA: AUC0-12h by dose (Section 3.3)

The paper reports the median (interquartile range) simulated AUC0-12h in
ng.h/mL. The model concentrations are converted from nmol/L to ng/mL with the
molecular weights above.

```{r pknca}
# The inverse-Gaussian input is vanishingly small just after the dose, and the
# integrator can return values a few ulps below zero there (of order 1e-17
# nmol/L). PKNCA returns NA for an AUC over a negative concentration, so the
# undershoot is first asserted to be pure numerical noise -- relative to each
# subject's peak -- and then floored at zero.
nca_input <- function(sim, conc_col, mw) {
  out <- sim |>
    dplyr::filter(!is.na(.data[[conc_col]])) |>
    dplyr::transmute(treatment, id, time, Cc = .data[[conc_col]] * mw / 1000) |>
    dplyr::group_by(id) |>
    dplyr::mutate(peak = max(Cc)) |>
    dplyr::ungroup()
  stopifnot(all(out$Cc >= -1e-9 * out$peak))
  out |>
    dplyr::mutate(Cc = pmax(Cc, 0)) |>
    dplyr::select(-peak)
}
run_nca <- function(conc_df, cohort) {
  conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | treatment + id,
                               concu = "ng/mL", timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(
    cohort |> dplyr::transmute(treatment, id, time = 0, dose = dose_mg),
    dose ~ time | treatment + id, doseu = "mg"
  )
  intervals <- data.frame(start = 0, end = 12, auclast = TRUE, cmax = TRUE, tmax = TRUE)
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}
nca_rosco <- run_nca(nca_input(dose_sim, "Cc", MW_ROSCO), dose_cohort)
nca_m3 <- run_nca(nca_input(dose_sim, "Cc_m3", MW_M3), dose_cohort)

published_rosco <- tibble::tribble(
  ~treatment, ~auclast,
  "200 mg", 477,
  "400 mg", 1436,
  "800 mg", 3711
)
published_m3 <- tibble::tribble(
  ~treatment, ~auclast,
  "200 mg", 1293,
  "400 mg", 2557,
  "800 mg", 4888
)
cmp <- dplyr::bind_rows(
  nlmixr2lib::ncaComparisonTable(nca_rosco, published_rosco, by = "treatment",
                                 params = "auclast", units = c(auclast = "ng*h/mL"),
                                 tolerance_pct = 20) |>
    dplyr::mutate(analyte = "Roscovitine"),
  nlmixr2lib::ncaComparisonTable(nca_m3, published_m3, by = "treatment",
                                 params = "auclast", units = c(auclast = "ng*h/mL"),
                                 tolerance_pct = 20) |>
    dplyr::mutate(analyte = "M3")
)
cmp |>
  dplyr::relocate(analyte) |>
  dplyr::rename("Analyte" = analyte, "Dose" = treatment) |>
  knitr::kable(caption = paste(
    "Simulated median AUC0-12h versus the Monte Carlo medians of Leven 2020",
    "Section 3.3. * differs by more than 20%."
  ))
```

```{r pknca_gate}
auc_tab <- function(nca, analyte) {
  as.data.frame(nca$result) |>
    dplyr::filter(PPTESTCD == "auclast") |>
    dplyr::select(treatment, id, auc = PPORRES) |>
    dplyr::mutate(analyte = analyte)
}
auc_all <- dplyr::bind_rows(auc_tab(nca_rosco, "rosco"), auc_tab(nca_m3, "m3"))
iqr_sim <- auc_all |>
  dplyr::group_by(analyte, treatment) |>
  dplyr::summarise(q25 = quantile(auc, 0.25), median = median(auc), q75 = quantile(auc, 0.75),
                   .groups = "drop")
iqr_pub <- tibble::tribble(
  ~analyte, ~treatment, ~pub_q25, ~pub_median, ~pub_q75,
  "rosco", "200 mg", 259, 477, 847,
  "rosco", "400 mg", 850, 1436, 2327,
  "rosco", "800 mg", 2335, 3711, 6329,
  "m3", "200 mg", 753, 1293, 2198,
  "m3", "400 mg", 1556, 2557, 4294,
  "m3", "800 mg", 2893, 4888, 8559
)
iqr_cmp <- dplyr::inner_join(iqr_sim, iqr_pub, by = c("analyte", "treatment"))
knitr::kable(iqr_cmp |>
               dplyr::rename("Analyte" = analyte, "Dose" = treatment,
                             "Sim. Q1" = q25, "Sim. median" = median, "Sim. Q3" = q75,
                             "Pub. Q1" = pub_q25, "Pub. median" = pub_median, "Pub. Q3" = pub_q75),
             digits = 0,
             caption = "Simulated versus published AUC0-12h quartiles (ng.h/mL).")
# Assert on the centre and on robust quantiles only. A 200-per-arm median of
# this highly variable absorption has a sampling error of several percent per
# arm, so the tight check is on the pooled centre across the six analyte-dose
# rows, and the per-row check only has to catch gross errors. A mis-transcribed
# volume, D50 or rate constant moves every row by tens of percent.
lr <- function(sim, pub) log(sim / pub)
stopifnot(
  abs(median(lr(iqr_cmp$median, iqr_cmp$pub_median))) < log(1.1),
  all(abs(lr(iqr_cmp$median, iqr_cmp$pub_median)) < log(1.3)),
  abs(median(lr(iqr_cmp$q25, iqr_cmp$pub_q25))) < log(1.15),
  abs(median(lr(iqr_cmp$q75, iqr_cmp$pub_q75))) < log(1.15)
)

# Dose proportionality (Figure 7 narrative): a fourfold dose increase raises
# the median roscovitine AUC0-12h well over fourfold (7.8-fold in the paper)
# and the median M3 AUC0-12h by less than the paper's 3.8-fold plus margin.
med <- function(a, d) iqr_sim$median[iqr_sim$analyte == a & iqr_sim$treatment == d]
stopifnot(
  med("rosco", "800 mg") / med("rosco", "200 mg") > 5,
  med("m3", "800 mg") / med("m3", "200 mg") < 4.5
)

# The M3 peak precedes the roscovitine peak in every dose arm.
tmax_med <- dplyr::bind_rows(
  as.data.frame(nca_rosco$result) |> dplyr::mutate(analyte = "rosco"),
  as.data.frame(nca_m3$result) |> dplyr::mutate(analyte = "m3")
) |>
  dplyr::filter(PPTESTCD == "tmax") |>
  dplyr::group_by(analyte, treatment) |>
  dplyr::summarise(tmax = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = analyte, values_from = tmax)
knitr::kable(tmax_med |> dplyr::rename("Dose" = treatment, "Median Tmax M3 (h)" = m3,
                                       "Median Tmax roscovitine (h)" = rosco),
             digits = 2)
stopifnot(all(tmax_med$rosco - tmax_med$m3 > 0.3))
```

# AUC ratio by dose (replicates Figure 8)

```{r figure8}
ratio <- auc_all |>
  tidyr::pivot_wider(names_from = analyte, values_from = auc) |>
  dplyr::mutate(ratio = rosco / m3 * MW_M3 / MW_ROSCO)
```

The ratio above is computed on molar AUCs (the ng.h/mL values multiplied back
by `MW_M3 / MW_ROSCO`), matching the paper's nmol/L modelling scale.

```{r figure8_plot}
ggplot(ratio, aes(treatment, ratio, fill = treatment)) +
  geom_violin(alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.size = 0.5, fill = "white") +
  labs(x = "Roscovitine dose", y = "Roscovitine / M3 AUC ratio", fill = "Dose",
       caption = "Replicates Figure 8 of Leven 2020.")

ratio_med <- ratio |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(median_ratio = median(ratio), .groups = "drop")
knitr::kable(ratio_med |> dplyr::rename("Dose" = treatment, "Median AUC ratio" = median_ratio),
             digits = 2)
```

The paper reports the median roscovitine/M3 AUC ratio rising from 0.37 at
200 mg to 0.79 at 800 mg (Section 3.3).

```{r figure8_gate}
r200 <- ratio_med$median_ratio[ratio_med$treatment == "200 mg"]
r800 <- ratio_med$median_ratio[ratio_med$treatment == "800 mg"]
stopifnot(abs(r200 / 0.37 - 1) < 0.2, abs(r800 / 0.79 - 1) < 0.2)
```

The ratio's median is taken on molar AUCs because the paper's own model works
in nmol/L; on ng.h/mL the ratio is smaller by the factor 354.45 / 368.43
(about 4%), which is within the tolerance either way.

# Effect of proton-pump inhibitors at 400 mg (replicates Figures 9 and 10)

```{r ppi}
ppi_cohort <- dplyr::bind_rows(
  make_cohort(n_per_arm, 400, ppi = 0L, id_offset = 5000L),
  make_cohort(n_per_arm, 400, ppi = 1L, id_offset = 6000L)
) |>
  dplyr::mutate(treatment = paste0("400 mg, PPI = ", CONMED_PPI))
ppi_sim <- rxode2::rxSolve(mod, build_events(ppi_cohort, obs_times),
                           returnType = "data.frame") |>
  dplyr::left_join(ppi_cohort |> dplyr::select(id, dose_mg, treatment), by = "id")

ppi_band <- ppi_sim |>
  dplyr::select(treatment, time, Roscovitine = Cc, `M3 metabolite` = Cc_m3) |>
  tidyr::pivot_longer(c(Roscovitine, `M3 metabolite`), names_to = "analyte", values_to = "conc") |>
  dplyr::group_by(treatment, analyte, time) |>
  dplyr::summarise(p10 = quantile(conc, 0.1), p50 = median(conc), p90 = quantile(conc, 0.9),
                   .groups = "drop") |>
  dplyr::mutate(analyte = factor(analyte, c("Roscovitine", "M3 metabolite")))

ggplot(ppi_band, aes(time, p50)) +
  geom_ribbon(aes(ymin = p10, ymax = p90), fill = "steelblue", alpha = 0.3) +
  geom_line(colour = "steelblue4") +
  facet_grid(analyte ~ treatment, scales = "free_y") +
  labs(x = "Time (h)", y = "Concentration (nmol/L)",
       caption = "Median and 10th-90th percentiles; replicates Figure 9 of Leven 2020.")
```

```{r ppi_nca}
nca_ppi_rosco <- run_nca(nca_input(ppi_sim, "Cc", MW_ROSCO), ppi_cohort)
nca_ppi_m3 <- run_nca(nca_input(ppi_sim, "Cc_m3", MW_M3), ppi_cohort)
ppi_res <- dplyr::bind_rows(
  as.data.frame(nca_ppi_rosco$result) |> dplyr::mutate(analyte = "rosco"),
  as.data.frame(nca_ppi_m3$result) |> dplyr::mutate(analyte = "m3")
) |>
  dplyr::filter(PPTESTCD %in% c("auclast", "tmax")) |>
  dplyr::select(analyte, treatment, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
ppi_sum <- ppi_res |>
  dplyr::group_by(analyte, treatment) |>
  dplyr::summarise(median_tmax = median(tmax), median_auc = median(auclast), .groups = "drop")
knitr::kable(ppi_sum |> dplyr::rename("Analyte" = analyte, "Group" = treatment,
                                      "Median Tmax (h)" = median_tmax,
                                      "Median AUC0-12h (ng.h/mL)" = median_auc),
             digits = 2)

ppi_ratio <- ppi_res |>
  dplyr::select(analyte, treatment, id, auclast) |>
  tidyr::pivot_wider(names_from = analyte, values_from = auclast) |>
  dplyr::mutate(ratio = rosco / m3 * MW_M3 / MW_ROSCO)
ggplot(ppi_ratio, aes(treatment, ratio, fill = treatment)) +
  geom_violin(alpha = 0.6) +
  geom_boxplot(width = 0.15, outlier.size = 0.5, fill = "white") +
  labs(x = NULL, y = "Roscovitine / M3 AUC ratio", fill = NULL,
       caption = "Replicates Figure 10 of Leven 2020.")
```

PPI use nearly doubles T1max (`exp(0.680)` = 1.97), which delays both the
roscovitine peak and -- because the pre-systemic M3 input shares T1max -- the
M3 peak, and flattens both, while leaving exposure essentially unchanged
(Figures 9 and 10: "no obvious effect on plasma exposure or AUCrosco/AUCM3
ratios").

```{r ppi_gate}
tm <- function(a, g) ppi_sum$median_tmax[ppi_sum$analyte == a & ppi_sum$treatment == g]
au <- function(a, g) ppi_sum$median_auc[ppi_sum$analyte == a & ppi_sum$treatment == g]
ratio_by_ppi <- ppi_ratio |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(m = median(ratio), .groups = "drop")
stopifnot(
  # Both peaks move later by a clear margin (about 1 h at typical values).
  tm("rosco", "400 mg, PPI = 1") - tm("rosco", "400 mg, PPI = 0") > 0.4,
  tm("m3", "400 mg, PPI = 1") - tm("m3", "400 mg, PPI = 0") > 0.4,
  # Exposure is essentially unchanged: magnitude, not sign.
  abs(log(au("rosco", "400 mg, PPI = 1") / au("rosco", "400 mg, PPI = 0"))) < log(1.3),
  abs(log(au("m3", "400 mg, PPI = 1") / au("m3", "400 mg, PPI = 0"))) < log(1.3),
  abs(log(ratio_by_ppi$m[2] / ratio_by_ppi$m[1])) < log(1.3)
)
```

# Sensitivity: cohort with the observed height range

```{r height_sensitivity}
ht_cohort <- dplyr::bind_rows(
  make_cohort(n_per_arm, 200, height = draw_height, id_offset = 7000L),
  make_cohort(n_per_arm, 400, height = draw_height, id_offset = 8000L),
  make_cohort(n_per_arm, 800, height = draw_height, id_offset = 9000L)
) |>
  dplyr::mutate(treatment = paste(dose_mg, "mg"))
ht_sim <- rxode2::rxSolve(mod, build_events(ht_cohort, obs_times),
                          returnType = "data.frame") |>
  dplyr::left_join(ht_cohort |> dplyr::select(id, dose_mg, treatment), by = "id")
ht_auc <- dplyr::bind_rows(
  auc_tab(run_nca(nca_input(ht_sim, "Cc", MW_ROSCO), ht_cohort), "rosco"),
  auc_tab(run_nca(nca_input(ht_sim, "Cc_m3", MW_M3), ht_cohort), "m3")
) |>
  dplyr::group_by(analyte, treatment) |>
  dplyr::summarise(median_observed_heights = median(auc), .groups = "drop")
ht_cmp <- iqr_cmp |>
  dplyr::select(analyte, treatment, median_170 = median, pub_median) |>
  dplyr::inner_join(ht_auc, by = c("analyte", "treatment"))
knitr::kable(ht_cmp |>
               dplyr::rename("Analyte" = analyte, "Dose" = treatment,
                             "Sim. median, 170 cm" = median_170,
                             "Published median" = pub_median,
                             "Sim. median, 158-182 cm cohort" = median_observed_heights),
             digits = 0,
             caption = "Median AUC0-12h (ng.h/mL) under the two height assumptions.")
```

At the observed median height the typical V/F is
`r round((166 / 170)^ini_vals[["e_ht_vc"]], 2)` times its reference value, so a
cohort centred on the observed heights is expected to show higher median
exposure than the reference-height simulation. Users simulating a real CF population should supply individual heights.

```{r height_factor}
v_factor <- (166 / 170)^ini_vals[["e_ht_vc"]]
stopifnot(abs(v_factor - 0.857) < 0.001)
```

# Assumptions and deviations

* **PPI effect on T1max, not dT2max.** The Results prose (Section 3.2) says
  "a substantial influence of PPI use on dT2max", but Table 3 and its footnote
  both place the coefficient on T1max. Figure 9 settles it: under PPI the M3
  peak moves later, which can only happen if the effect is on T1max (the M3
  input shares T1max; dT2max affects the second roscovitine input only). The
  model follows Table 3.
* **Covariate forms.** Equations 9 and 10 give `ln(theta_i) = ln(theta_pop) +
  beta * CAT` for PPI and `ln(theta_i) = ln(theta_pop) + beta * ln(COV / COVref)`
  for height, with COVref = 170 cm (Section 3.2). The height exponent 6.47 is
  large but acts over the narrow observed range: V/F spans 0.62-1.55 times
  its typical value between 158 and 182 cm.
* **Correlation signs.** The T1max-CV2 correlation is -0.839 (negative) and the
  dT2max-CV1 correlation +0.624, as printed in the Table 3 PDF layout.
* **No IIV on pi1, k12, k21 or kmet**, and pi1 is held on the logit scale
  (logit-normal, Section 2.5) with no random effect, per Table 3.
* **Virtual-cohort covariates.** The paper does not say which covariate values
  its Monte Carlo simulations used. The primary cohort uses the reference
  height (170 cm) and the observed PPI prevalence, which reproduces the
  published AUC medians and the AUC-ratio medians; a cohort spanning the
  observed heights is shown as a sensitivity analysis.
* **Dose units.** The model is dosed in umol of roscovitine; convert a
  milligram dose with `mg / 354.45 * 1000`.
* **Single-dose model.** The model was developed from first-dose data only.
  Repeated dosing restarts the input clock at each dose via `tad()` and uses
  that dose's amount via `podo()`; this is appropriate for once-daily dosing
  but does not describe the inter-occasion variability the authors anticipate
  (Discussion).
