---
title: "Vancomycin in infants by renal function (Li 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vancomycin in infants by renal function (Li 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Li 2021 developed three population PK models of vancomycin in parallel from
one retrospective cohort of Chinese infants aged 1-24 months. The cohort was
split by renal function (modified-Schwartz eGFR):

| Paper label | nlmixr2lib model | Group | Covariates |
|-------------|------------------|-------|------------|
| Model 1 | `Li_2021_vancomycin_normal` | normal renal function, eGFR 30-86 mL/min/1.73 m^2 | WT on CL and V; SCR on CL |
| Model 2 | `Li_2021_vancomycin_arc` | augmented renal clearance (ARC), eGFR >= 86 mL/min/1.73 m^2 | WT on CL and V |
| Model 3 | `Li_2021_vancomycin_all` | all infants with eGFR >= 30 mL/min/1.73 m^2 (pooled) | WT on CL and V; SCR on CL |

* Citation: `r rxode2::rxode(readModelDb("Li_2021_vancomycin_normal"))$reference`
* Article: [Front Pediatr 2021;9:713588](https://doi.org/10.3389/fped.2021.713588)

Descriptions:

* Model 1: `r rxode2::rxode(readModelDb("Li_2021_vancomycin_normal"))$description`
* Model 2: `r rxode2::rxode(readModelDb("Li_2021_vancomycin_arc"))$description`
* Model 3: `r rxode2::rxode(readModelDb("Li_2021_vancomycin_all"))$description`

# Population

Infants aged 1-24 months who received IV vancomycin for at least 3 days at the
Children's Hospital of Nanjing Medical University between January 2017 and
July 2021 were enrolled (Li 2021 Methods "Patients"). Each had at least one
trough and one peak concentration, and an eGFR of at least 30
mL/min/1.73 m^2. Vancomycin was given as an IV infusion lasting at least
60 min, two to four times daily, with doses adjusted by therapeutic drug
monitoring. Whole-blood samples were drawn 30 min before the fifth dose
(trough) and 30 min after the fifth dose (peak), and assayed by EMIT.

Li 2021 Table 2 summarises the three modelling groups:

| | Model 1 (normal) | Model 2 (ARC) | Model 3 (all) |
|---|---|---|---|
| Patients (male/female) | 61 (37/24) | 64 (41/23) | 115 (73/42) |
| Concentrations (trough/peak) | 135 (69/66) | 139 (88/51) | 276 (158/118) |
| Preterm infants | 29 | 17 | 46 |
| Weight, kg, median (range) | 2.25 (1.15-13) | 4.60 (2.2-14) | 3.3 (1.15-14) |
| SCR, umol/L, median (range) | 27.1 (14-315) | 16.8 (8-29) | 18.95 (8-147) |
| eGFR, mL/min/1.73 m^2, median (range) | 57.56 (30-85.56) | 128 (90.8-280) | 98.7 (30-280) |
| Daily dose, mg/day, median (range) | 75 (24-320) | 216 (50-640) | 145 (24-640) |
| Trough, mg/L, median (range) | 7.05 (2.1-22.2) | 7 (2.1-23) | 7.1 (2.1-23) |
| Peak, mg/L, median (range) | 19.9 (6.8-46.9) | 18.7 (6.1-36.6) | 19.1 (6.1-46.9) |

The Model 1 and Model 2 group sizes add up to more than 115. Patients whose
renal function changed during treatment were counted in both groups (Results
"Patients"). The same information is available programmatically, for example
`readModelDb("Li_2021_vancomycin_normal")()$population`.

# Source trace

Each model has the same structure: a one-compartment model with first-order
elimination (NONMEM ADVAN1 TRANS2), exponential IIV on CL only, and a
proportional residual error. The covariate equations are printed both as the
header row of Tables 5-7 and as the three final-model equations in Results
"Model Building":

* Model 1: `CL = theta1 * (WT/2.25)^theta3 * exp(theta5 * SCR/27.1)`; `V = theta2 * (WT/2.25)^theta4`
* Model 2: `CL = theta1 * (WT/4.6)^theta3`; `V = theta2 * (WT/4.6)^theta4`
* Model 3: `CL = theta1 * (WT/3.45)^theta3 * exp(theta5 * SCR/19)`; `V = theta2 * (WT/3.45)^theta4`

| Parameter | Model 1 | Model 2 | Model 3 | Source |
|-----------|---------|---------|---------|--------|
| `lcl` (theta1, L/h) | log(0.407) | log(0.756) | log(0.707) | Tables 5 / 6 / 7 |
| `lvc` (theta2, L) | log(1.86) | log(4.89) | log(3.39) | Tables 5 / 6 / 7 |
| `e_wt_cl` (theta3) | 1.24 | 1.03 | 1.23 | Tables 5 / 6 / 7 |
| `e_wt_vc` (theta4) | 1.28 | 0.918 | 1.29 | Tables 5 / 6 / 7 |
| `e_creat_cl` (theta5) | -0.533 | n/a | -0.377 | Tables 5 / 7 |
| `etalcl` (BSV_CL squared) | 0.315^2 | 0.312^2 | 0.311^2 | Tables 5 / 6 / 7 |
| `propSd` (PROP_RV) | 0.319 | 0.319 | 0.335 | Tables 5 / 6 / 7 |
| WT reference (kg) | 2.25 | 4.6 | 3.45 | Table header equations; Results equations |
| SCR divisor (umol/L) | 27.1 | n/a | 19 | Table header equations; Results equations |
| 1-cmt, first-order elimination | | | | Methods "Base Model" |
| Exponential IIV | | | | Methods "Base Model" |
| Proportional residual `Y = F*(1+eps)` | | | | Methods "Base Model"; PROP_RV row |

## Scale of `BSV_CL` and `PROP_RV`

Li 2021 does not say whether `BSV_CL` and `PROP_RV` are standard deviations or
variances. The footnotes call `PROP_RV` a "proportional residual variance",
which suggests a variance. The observed peak concentrations rule that reading
out.

A peak sampled 30 min after the infusion carries the full proportional
residual error. The coefficient of variation (CV) of observed peaks therefore
cannot be smaller than the residual SD. If `PROP_RV` were a variance, the
residual SD would be `sqrt(0.319) = 0.565`. Yet the peaks in every group of
Table 2 vary with a CV of only 36-39%, even though that CV also includes
between-subject, covariate and dose variation. So `PROP_RV` must be an SD.
The same table reports `BSV_CL` on the same scale, so it is also read as the
SD of eta, and its square gives the variance used in `ini()`.

```{r scale-check}
scale_check <- tibble::tribble(
  ~model,    ~prop_rv, ~peak_mean, ~peak_sd,
  "Model 1", 0.319,    22.28,      8.75,
  "Model 2", 0.319,    18.82,      6.76,
  "Model 3", 0.335,    20.61,      7.67
) |>
  mutate(
    observed_peak_cv = peak_sd / peak_mean,
    residual_sd_if_variance = sqrt(prop_rv),
    residual_sd_if_sd = prop_rv
  )

# Table 2 arithmetic, no simulation. The variance reading would need an
# observed peak CV of at least sqrt(PROP_RV) = 0.565-0.579; observed is
# 0.36-0.39. The SD reading (0.319-0.335) sits below it, as it must.
stopifnot(
  all(scale_check$observed_peak_cv < scale_check$residual_sd_if_variance),
  all(scale_check$observed_peak_cv > scale_check$residual_sd_if_sd)
)

scale_check |>
  dplyr::rename(
    "Model" = model,
    "PROP_RV (Tables 5-7)" = prop_rv,
    "Observed peak CV (Table 2)" = observed_peak_cv,
    "Residual SD if variance" = residual_sd_if_variance,
    "Residual SD if SD" = residual_sd_if_sd
  ) |>
  dplyr::select(-peak_mean, -peak_sd) |>
  knitr::kable(digits = 3, caption = "Observed peak variability versus the two readings of PROP_RV.")
```

# Typical-value check against the observed troughs and peaks

Li 2021 reports no NCA table, and its VPCs (Figures 4-6) are plotted against
time since the first dose, pooled across dose levels. The most direct check
of the printed equations uses the observed median troughs and peaks in
Table 2. The typical infant of each group is given the Table 2 median weight
and SCR. That infant receives the Table 2 median daily dose as 1-h infusions
every 8 h. The trough is read 30 min before the fifth dose and the peak 30 min
after the fifth infusion ends, which matches the sampling design.

These are typical-value predictions compared with population medians, not
identical quantities. The check is still sharp on the covariate equations.
The creatinine factor is not centred: at the reference SCR, `exp(theta5)` is
0.59 for Model 1 and 0.69 for Model 3. Dropping it, or re-centring it as
`exp(theta5 * (SCR/27.1 - 1))`, puts the Model 1 typical trough 50% and its
peak 34% below the observed medians, so both gates below would fail.

```{r typical-check}
typical_cases <- tibble::tribble(
  ~model,                        ~label,    ~WT,  ~CREAT, ~daily_dose, ~obs_trough, ~obs_peak,
  "Li_2021_vancomycin_normal",   "Model 1", 2.25, 27.1,   75,          7.05,        19.9,
  "Li_2021_vancomycin_arc",      "Model 2", 4.60, 16.8,   216,         7.0,         18.7,
  "Li_2021_vancomycin_all",      "Model 3", 3.3,  18.95,  145,         7.1,         19.1
)

tau <- 8
tinf <- 1
n_dose <- 5
dose_times <- seq(0, by = tau, length.out = n_dose)
t_trough <- dose_times[n_dose] - 0.5
t_peak <- dose_times[n_dose] + tinf + 0.5

typical_one <- function(i) {
  case <- typical_cases[i, ]
  amt <- case$daily_dose * tau / 24
  ev <- dplyr::bind_rows(
    tibble(time = dose_times, evid = 1L, amt = amt, rate = amt / tinf),
    tibble(time = c(t_trough, t_peak), evid = 0L, amt = 0, rate = 0)
  ) |>
    mutate(id = 1L, cmt = "central", WT = case$WT, CREAT = case$CREAT) |>
    arrange(time, desc(evid))
  mod_typ <- rxode2::zeroRe(readModelDb(case$model))
  s <- as.data.frame(rxode2::rxSolve(mod_typ, events = ev, rtol = 1e-10, atol = 1e-12))
  tibble(
    label = case$label,
    pred_trough = s$Cc[s$time == t_trough],
    pred_peak = s$Cc[s$time == t_peak]
  )
}

typical_res <- dplyr::bind_rows(lapply(seq_len(nrow(typical_cases)), typical_one)) |>
  left_join(dplyr::select(typical_cases, label, obs_trough, obs_peak), by = "label") |>
  mutate(
    trough_pct = 100 * (pred_trough / obs_trough - 1),
    peak_pct = 100 * (pred_peak / obs_peak - 1)
  )

# Deterministic (zeroRe, one subject per model). Measured: trough +19 / -0 /
# +13 %, peak -8 / -5 / +0 % for Models 1 / 2 / 3. A dropped or re-centred
# creatinine factor gives -50 % (trough) and -34 % (peak) for Model 1; a dose
# or volume mis-transcription moves the peaks by tens of percent.
stopifnot(
  nrow(typical_res) == 3L,
  all(abs(typical_res$trough_pct) < 30),
  all(abs(typical_res$peak_pct) < 20)
)

typical_res |>
  dplyr::rename(
    "Model" = label,
    "Predicted trough (mg/L)" = pred_trough,
    "Observed median trough (mg/L)" = obs_trough,
    "Trough difference (%)" = trough_pct,
    "Predicted peak (mg/L)" = pred_peak,
    "Observed median peak (mg/L)" = obs_peak,
    "Peak difference (%)" = peak_pct
  ) |>
  knitr::kable(digits = 2, caption = "Typical-infant trough and peak around the fifth dose (median daily dose, q8h, 1-h infusion) versus the Table 2 observed medians.")
```

# Virtual cohort

The original data are not public. Each model gets its own virtual cohort of
200 infants, drawn to resemble its Table 2 group. Weight and SCR are
log-normal. The median is the Table 2 median, and the log-SD is set so the
distribution also reproduces the Table 2 mean (`sdlog = sqrt(2 * log(mean /
median))`). Values outside the Table 2 range are redrawn. Every infant gets
the group's median daily dose per kg (Table 2 median daily dose divided by
median weight) as 1-h infusions every 8 h, for 12 doses.

```{r cohort}
rxode2::rxSetSeed(20210920)
set.seed(20210920)

n_per_arm <- 200L
n_dose_sim <- 12L
dose_times_sim <- seq(0, by = tau, length.out = n_dose_sim)

# Reject-and-redraw inside [lo, hi]; clamping would pile subjects on the bounds.
rlnorm_range <- function(n, median, mean, lo, hi) {
  sdlog <- sqrt(2 * log(mean / median))
  out <- numeric(0)
  while (length(out) < n) {
    x <- stats::rlnorm(4 * n, log(median), sdlog)
    out <- c(out, x[x >= lo & x <= hi])
  }
  out[seq_len(n)]
}

arms <- tibble::tribble(
  ~model,                      ~label,    ~wt_med, ~wt_mean, ~wt_lo, ~wt_hi, ~cr_med, ~cr_mean, ~cr_lo, ~cr_hi, ~dd_med,
  "Li_2021_vancomycin_normal", "Model 1", 2.25,    2.86,     1.15,   13,     27.1,    30.80,    14,     315,    75,
  "Li_2021_vancomycin_arc",    "Model 2", 4.60,    5.34,     2.2,    14,     16.8,    16.84,    8,      29,     216,
  "Li_2021_vancomycin_all",    "Model 3", 3.3,     4.11,     1.15,   14,     18.95,   23.54,    8,      147,    145
)

obs_times <- sort(unique(c(
  seq(0, max(dose_times_sim) + tau, by = 0.5),
  t_trough, t_peak
)))

make_arm <- function(i) {
  a <- arms[i, ]
  subj <- tibble(
    id = (i - 1L) * n_per_arm + seq_len(n_per_arm),
    WT = rlnorm_range(n_per_arm, a$wt_med, a$wt_mean, a$wt_lo, a$wt_hi),
    # Model 2 has no creatinine term; SCR is drawn for every arm so the
    # event tables share one shape (its Table 2 mean ~ median gives a
    # near-degenerate spread, which is harmless there).
    CREAT = rlnorm_range(n_per_arm, a$cr_med, max(a$cr_mean, a$cr_med * 1.001), a$cr_lo, a$cr_hi),
    mgkg_day = a$dd_med / a$wt_med
  ) |>
    mutate(amt = mgkg_day * WT * tau / 24)
  doses <- tidyr::expand_grid(id = subj$id, time = dose_times_sim) |>
    left_join(dplyr::select(subj, id, amt), by = "id") |>
    mutate(evid = 1L, rate = amt / tinf)
  obs <- tidyr::expand_grid(id = subj$id, time = obs_times) |>
    mutate(evid = 0L, amt = 0, rate = 0)
  dplyr::bind_rows(doses, obs) |>
    left_join(dplyr::select(subj, id, WT, CREAT), by = "id") |>
    mutate(cmt = "central", model = a$model, label = a$label) |>
    arrange(id, time, desc(evid))
}

events_all <- dplyr::bind_rows(lapply(seq_len(nrow(arms)), make_arm))
stopifnot(!anyDuplicated(unique(events_all[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
sim <- dplyr::bind_rows(lapply(seq_len(nrow(arms)), function(i) {
  ev <- dplyr::filter(events_all, model == arms$model[i])
  rxode2::rxSolve(
    readModelDb(arms$model[i]),
    events = ev,
    keep = c("label", "WT", "CREAT")
  ) |>
    as.data.frame()
}))
```

# Replicate published figures

Figures 4-6 of Li 2021 are VPCs of observed concentrations against time since
the first dose. The patients were sampled around their fifth dose, and their
treatment courses ran for different lengths. The plot below shows the same
quantity for the virtual cohorts: the median and 90% prediction interval of
simulated whole-blood concentrations (with residual error). The dashed lines
mark the Table 2 observed median trough and peak.

```{r figure-vpc, fig.width = 7, fig.height = 6}
obs_lines <- tibble::tribble(
  ~label,    ~trough, ~peak,
  "Model 1", 7.05,    19.9,
  "Model 2", 7.0,     18.7,
  "Model 3", 7.1,     19.1
) |>
  tidyr::pivot_longer(c(trough, peak), names_to = "sample", values_to = "conc")

sim |>
  group_by(label, time) |>
  summarise(
    Q05 = quantile(sim, 0.05, na.rm = TRUE),
    Q50 = quantile(sim, 0.50, na.rm = TRUE),
    Q95 = quantile(sim, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "firebrick") +
  geom_hline(data = obs_lines, aes(yintercept = conc, linetype = sample)) +
  facet_wrap(~label, ncol = 1) +
  labs(
    x = "Time since first dose (h)", y = "Vancomycin (mg/L)", linetype = "Table 2 median",
    title = "Simulated concentrations, median and 90% PI",
    caption = "Compare with Figures 4-6 of Li 2021 (VPCs of Models 1-3)."
  )
```

Simulated troughs and peaks around the fifth dose, with residual error,
compared with the Table 2 observed medians and ranges. The simulated CVs are
larger than the observed ones. In the study, doses were adjusted by drug
monitoring, which narrows the observed spread, and the virtual cohort draws
some infants from the high-creatinine tail. The comparison does not favour
the variance reading of `BSV_CL` and `PROP_RV`. That reading would widen the
simulated spread further.

```{r fifth-dose-summary}
fifth <- sim |>
  filter(time %in% c(t_trough, t_peak)) |>
  mutate(sample = ifelse(time == t_trough, "trough", "peak")) |>
  group_by(label, sample) |>
  summarise(
    sim_median = median(sim),
    sim_p05 = quantile(sim, 0.05),
    sim_p95 = quantile(sim, 0.95),
    sim_cv = sd(sim) / mean(sim),
    .groups = "drop"
  ) |>
  left_join(
    tibble::tribble(
      ~label,    ~sample,  ~obs_median, ~obs_range,  ~obs_cv,
      "Model 1", "trough", 7.05,        "2.1-22.2",  4.67 / 8.29,
      "Model 1", "peak",   19.9,        "6.8-46.9",  8.75 / 22.28,
      "Model 2", "trough", 7.0,         "2.1-23",    4.66 / 7.93,
      "Model 2", "peak",   18.7,        "6.1-36.6",  6.76 / 18.82,
      "Model 3", "trough", 7.1,         "2.1-23",    4.75 / 8.31,
      "Model 3", "peak",   19.1,        "6.1-46.9",  7.67 / 20.61
    ),
    by = c("label", "sample")
  )

# Cohort medians against observed medians. The observed doses were TDM-adjusted
# and varied per infant, so the check is on the centre with a wide envelope.
# Measured simulated/observed median ratios: 0.89-1.27 across the six cells.
# The deterministic typical-value check above is the sharp structural gate;
# this one catches unit or dose errors, which move the medians several-fold.
stopifnot(
  nrow(fifth) == 6L,
  all(fifth$sim_median / fifth$obs_median > 0.6),
  all(fifth$sim_median / fifth$obs_median < 1.6)
)

fifth |>
  dplyr::rename(
    "Model" = label, "Sample" = sample,
    "Simulated median" = sim_median, "Simulated 5th pct" = sim_p05,
    "Simulated 95th pct" = sim_p95, "Simulated CV" = sim_cv,
    "Observed median" = obs_median, "Observed range" = obs_range,
    "Observed CV" = obs_cv
  ) |>
  knitr::kable(digits = 2, caption = "Troughs (30 min before the fifth dose) and peaks (30 min after the fifth infusion), simulated versus Li 2021 Table 2 (mg/L).")
```

# PKNCA validation

Steady-state NCA over the 12th dosing interval, grouped by model. At steady
state, the AUC over one dosing interval equals the dose divided by the
individual CL. The check below compares the two for each virtual infant.

```{r pknca}
t_start <- max(dose_times_sim)
t_end <- t_start + tau

sim_nca <- sim |>
  filter(!is.na(Cc), time >= t_start, time <= t_end) |>
  mutate(Cc = pmax(Cc, 0)) |>
  dplyr::select(id, time, Cc, label)

dose_nca <- events_all |>
  filter(evid == 1, time == t_start) |>
  dplyr::select(id, time, amt, label)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | label + id, concu = "mg/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_nca, amt ~ time | label + id, doseu = "mg")
intervals <- data.frame(
  start = t_start, end = t_end,
  cmax = TRUE, cmin = TRUE, tmax = TRUE, auclast = TRUE, cav = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

knitr::kable(summary(nca_res), caption = "Simulated steady-state NCA (12th dosing interval, q8h) by model.")

auc_check <- as.data.frame(nca_res) |>
  filter(PPTESTCD == "auclast") |>
  dplyr::select(id, label, auclast = PPORRES) |>
  left_join(
    sim |> filter(time == t_start) |> dplyr::select(id, cl),
    by = "id"
  ) |>
  left_join(dplyr::select(dose_nca, id, amt), by = "id") |>
  mutate(pct_diff = 100 * (auclast * cl / amt - 1))

# AUCtau * CL_i / Dose = 1 at steady state. Deviations come from the
# 0.5-h trapezoid grid and from the slowest-clearing infants not being fully
# at steady state after 88 h. Measured: median -0.04 %, 90th percentile of
# |difference| 0.09-0.21 % per model; the bounds leave room for a cohort with
# more slow-clearing infants while still failing on any dose or CL error.
stopifnot(
  nrow(auc_check) == 3L * n_per_arm,
  abs(median(auc_check$pct_diff)) < 1,
  quantile(abs(auc_check$pct_diff), 0.9) < 5
)

auc_check |>
  group_by(label) |>
  summarise(
    median_pct_diff = median(pct_diff),
    p90_abs_pct_diff = quantile(abs(pct_diff), 0.9),
    .groups = "drop"
  ) |>
  dplyr::rename(
    "Model" = label,
    "Median (AUCtau*CL/Dose - 1), %" = median_pct_diff,
    "90th pct of |difference|, %" = p90_abs_pct_diff
  ) |>
  knitr::kable(digits = 2, caption = "Steady-state AUCtau against Dose / individual CL.")
```

Li 2021 reports no NCA parameters, so there is no published NCA table to
compare against. The comparison with the paper is the trough and peak check
above.

# Assumptions and deviations

- **IIV and residual-error scale.** Tables 5-7 do not say whether `BSV_CL`
  and `PROP_RV` are SDs or variances, and the footnote calls `PROP_RV` a
  "variance". The maintainers read both as SDs. As shown in the "Scale of
  `BSV_CL` and `PROP_RV`" section, the variance reading needs a residual SD
  of 0.565-0.579, which is larger than the observed peak CV of 0.36-0.39 in
  every group. `etalcl` is therefore `BSV_CL^2` and `propSd` is `PROP_RV`.
- **Uncentred creatinine factor.** The creatinine term is encoded exactly as
  printed, `exp(theta5 * SCR / 27.1)` (Model 1) and `exp(theta5 * SCR / 19)`
  (Model 3), in both the Table 5/7 header equations and the Results
  equations. It equals `exp(theta5)`, not 1, at the reference SCR, so
  `exp(lcl)` is not the typical CL of the reference infant. The as-printed
  form is the one that reproduces the observed troughs and peaks (typical-value
  check above).
- **Model 3 weight reference.** The Model 3 equations divide weight by
  3.45 kg, while Table 2 gives a median of 3.3 kg for that group. The printed
  equation value (3.45 kg) is used. The Model 3 SCR divisor (19 umol/L)
  matches the Table 2 median of 18.95 umol/L after rounding.
- **Group labels.** Table 2's column headers call the Model 1 group "reduced
  renal function" and the Model 2 group "normal renal function". The
  abstract, Methods and Results call them the normal-renal-function and ARC
  groups. The model names follow the abstract, Methods and Results.
- **Dosing in the vignette.** The paper gives the daily dose (Table 2) and
  says infusions last at least 60 min and are given two to four times daily.
  The vignette uses 1-h infusions every 8 h at each group's median mg/kg/day.
  "30 min after the fifth administration" is read as 30 min after the end of
  the fifth infusion.
- **Covariate distributions.** Weight and SCR are drawn independently. The
  paper does not report their correlation.
- **Covariates not retained.** Sex, age, height, ALT, AST, BUN, cystatin C,
  albumin, total protein, eGFR, and meropenem or imipenem co-administration
  were screened (Table 4 for Model 1) but are not in any final model. eGFR
  defines which model applies (Model 1: 30-86; Model 2: >= 86
  mL/min/1.73 m^2) but is not a model covariate.
- **Specimen.** Li 2021 assayed whole blood (Methods "Bioassay"), so the
  concentrations are whole-blood vancomycin concentrations.
- **Errata.** No correction notice for this article was found on Crossref as
  of 2026-09-29.
