---
title: "Ceftazidime (Li 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ceftazidime (Li 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 6)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Li_2024_ceftazidime"))$reference`
* Description: `r rxode2::rxode(readModelDb("Li_2024_ceftazidime"))$description`
* Article: <https://doi.org/10.3389/fphar.2024.1470350>

Li 2024 developed a population PK model for intravenous ceftazidime in
critically ill children admitted to the paediatric intensive care unit (PICU)
of Wuhan Children's Hospital, and used it to recommend weight- and
renal-function-stratified dosing regimens that reach a pharmacodynamic target
of 70% *f*T>MIC.

# Population

The analysis pooled 100 ceftazidime serum concentrations from 88 critically ill
children (Li 2024 Table 1). Ages ranged from 0.03 to 15 years (median 5.17,
mean 5.43, SD 4.10) and body weights from 2.80 to 95 kg (median 18.35, mean
21.07, SD 14.70); 56 subjects (63.6%) were male. Renal function was estimated
with the modified Schwartz equation and spanned 40.42 to 197.15
mL/min/1.73 m^2 (median 116.93). The cohort skews strongly toward *augmented*
renal clearance rather than impairment: 43 children had eGFR 120-200, 31 had
90-120, 9 had 60-90 and only 5 had 30-60 mL/min/1.73 m^2. Ceftazidime was given
intravenously at 25-100 mg/kg with a median loading dose of 48.38 mg/kg
(range 21.05-76), and serum was assayed by validated HPLC-UV over a linear
range of 0.025-100 ug/mL.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Li_2024_ceftazidime")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Li_2024_ceftazidime.R`. The table below
collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL at the reference subject) | 7.76 L/h | Li 2024 Table 3, RSE 10.40%, bootstrap 95% CI 6.14-10.03 |
| `lvc` (Vd at the reference subject) | 27.83 L | Li 2024 Table 3, RSE 7.85%, bootstrap 95% CI 21.04-37.70 |
| `e_wt_vc` (exponent on WT for Vd) | 1, fixed | Li 2024 Table 3 theta1, reported "1 (fixed)"; footnote maps theta1 to weight-on-V |
| `e_wt_cl` (exponent on WT for CL) | 0.90 | Li 2024 Table 3 theta2, RSE 6.41% |
| `e_crcl_cl` (exponent on eGFR for CL) | 0.38 | Li 2024 Table 3 theta3, RSE 33.59% |
| `etalcl` (IIV variance on CL) | 0.06 | Li 2024 Table 3 omega^2_CL, RSE 21.40% |
| `etalvc` (IIV variance on Vd) | 0.04 | Li 2024 Table 3 omega^2_Vd, RSE 39.77% |
| `addSd` (additive residual error) | 1.20 mg/L | Li 2024 Table 3 sigma, RSE 14.60%; Sect. 3.2 states the additive model beat the proportional and combined models |
| Reference weight | 70 kg | Li 2024 Eqs. 5 and 6 denominators; Results text under Table 3 |
| Reference eGFR | 116.93 mL/min/1.73 m^2 | Li 2024 Eq. 5 denominator; Results text under Table 3 ("a median eGFR of 116.93") |
| `cl <- ... * (WT/70)^0.90 * (CRCL/116.93)^0.38` | n/a | Li 2024 Equation 5 |
| `vc <- ... * (WT/70)^1` | n/a | Li 2024 Equation 6 |
| `d/dt(central) <- -kel * central` | n/a | Li 2024 Sect. 3.2: "one-compartment model with linear elimination ... parameterized by Vd and CL" |
| Exponential IIV, `P_i = theta * exp(eta_i)` | n/a | Li 2024 Equation 1 |

Equations 5 and 6 are typeset as stacked fractions in the published PDF and are
dropped by markdown converters (they appear as `formula-not-decoded`). They were
recovered with `pdftotext -layout`, which resolves both in full:

```
CL (L/h) = 7.76  * (weight/70)^0.9 * (eGFR/116.93)^0.38 * exp(eta_CL)
Vd (L)   = 27.83 * (weight/70)                          * exp(eta_Vd)
```

Note that Eq. 6 carries no printed exponent on `(weight/70)`, which is
consistent with Table 3 reporting theta1 as exactly 1 (fixed).

# Structural checks

Before simulating anything, three closed-form consequences of the published
parameters can be checked against statements the paper makes elsewhere. These
are deterministic identities, so they are asserted exactly.

```{r structural}
ref_wt   <- 70
ref_crcl <- 116.93
cl_ref   <- 7.76
vd_ref   <- 27.83

# Reference-subject elimination half-life.
t_half_ref <- log(2) * vd_ref / cl_ref

# kel(WT, eGFR) = (CL/Vd) * (WT/70)^(0.90 - 1) * (eGFR/116.93)^0.38.
# The weight exponent pair (0.90 on CL, 1 on Vd) therefore makes kel scale as
# WT^-0.10: SMALLER children eliminate faster per unit volume.
kel_fun <- function(wt, crcl) {
  (cl_ref / vd_ref) * (wt / ref_wt)^(0.90 - 1) * (crcl / ref_crcl)^0.38
}

# A mg/kg dose given as an IV bolus produces a WEIGHT-INDEPENDENT peak, because
# Vd scales with WT^1 exactly: C0 = (d * WT) / (27.83 * WT / 70) = d * 70/27.83.
c0_per_mgkg <- ref_wt / vd_ref

data.frame(
  quantity = c("t1/2 at WT=70 kg, eGFR=116.93 (h)",
               "kel exponent on weight",
               "C0 per 1 mg/kg IV bolus (mg/L)"),
  value    = c(round(t_half_ref, 3), 0.90 - 1, round(c0_per_mgkg, 4))
) |>
  knitr::kable(caption = "Closed-form consequences of the Li 2024 Table 3 estimates.")

stopifnot(
  # Li 2024 Introduction: "Ceftazidime has a half-life of 1.5-2.5 h in patients
  # with normal renal function". The reference subject sits inside that window.
  t_half_ref > 1.5, t_half_ref < 2.5,
  # kel falls with weight, which is the mechanism behind the Li 2024 Sect. 3.4
  # observation that concentrations are LOWER in children under 10 kg.
  kel_fun(5, ref_crcl) > kel_fun(60, ref_crcl),
  # Renal function raises kel.
  kel_fun(20, 160) > kel_fun(20, 45)
)
```

The reference half-life of `r round(t_half_ref, 2)` h lands inside the
1.5-2.5 h range the paper quotes for normal renal function, and the weight
exponent on `kel` is negative, reproducing the direction of the paper's own
Section 3.4 finding.

# Virtual cohorts

Original observed data are not publicly available. Two virtual populations are
used below: a typical-value grid for the Figure 5 replication, and a
covariate-sampled cohort for the target-attainment check.

```{r cohort-setup}
# set.seed() seeds R's RNG. It does NOT seed rxode2's simulation RNG, and
# rxode2's streams are partitioned PER SOLVER THREAD -- so a cohort drawn with
# IIV is reproducible on this machine and different on a machine with a
# different thread count. Every assertion on a cohort-derived quantity below is
# written to hold for ANY cohort the model can produce (see pattern 12 of
# references/known-vignette-failure-patterns.md).
set.seed(20241127)
rxode2::rxSetSeed(20241127)

mod <- readModelDb("Li_2024_ceftazidime")

# Li 2024 Sect. 3.4 stratifies by four weight bands and four eGFR bands.
wt_bands <- tibble::tribble(
  ~wt_band,  ~wt_lo, ~wt_hi,
  "<10 kg",     2.8,     10,   # lower edge is the cohort minimum (Table 1)
  "10-30 kg",    10,     30,
  "30-50 kg",    30,     50,
  "50-70 kg",    50,     70
)
egfr_bands <- tibble::tribble(
  ~egfr_band,                ~egfr_lo, ~egfr_hi,
  "30-60 (moderate)",              30,       60,
  "60-90 (mild)",                  60,       90,
  "90-120 (normal)",               90,      120,
  "120-200 (augmented)",          120,      200
)

# Band midpoints for the typical-value figure.
wt_bands$wt_mid     <- (wt_bands$wt_lo + wt_bands$wt_hi) / 2
egfr_bands$egfr_mid <- (egfr_bands$egfr_lo + egfr_bands$egfr_hi) / 2

wt_bands$wt_band     <- factor(wt_bands$wt_band, levels = wt_bands$wt_band)
egfr_bands$egfr_band <- factor(egfr_bands$egfr_band, levels = egfr_bands$egfr_band)
```

# Replicate Figure 5

Li 2024 Figure 5 simulates every weight x renal-function subgroup on a
standardised 30 mg/kg every-12-hour regimen over one week, and shows a single
smooth profile per panel -- a typical-value prediction rather than a VPC. It is
reproduced here with the random effects zeroed.

```{r figure-5, fig.height = 6.5}
fig5_dose_mgkg <- 30   # Li 2024 Sect. 3.4: "a standardized dose of 30 mg/kg every 12 h"
fig5_tau       <- 12
fig5_end       <- 168  # Li 2024 Figure 5 x-axis runs 0-168 h

fig5_subj <- tidyr::crossing(
  wt_bands   |> dplyr::select(wt_band, WT = wt_mid),
  egfr_bands |> dplyr::select(egfr_band, CRCL = egfr_mid)
) |>
  dplyr::mutate(id = dplyr::row_number(), amt_mg = fig5_dose_mgkg * WT)

# Dosing rows: IV bolus into `central`. Li 2024 does not report an infusion
# duration anywhere (Sect. 2.2 says only "administered intravenously"); the
# peak heights in Figure 5 are consistent with effectively instantaneous input
# (see the assertion below). Recorded as an assumption in the Errata.
fig5_dose <- fig5_subj |>
  tidyr::crossing(time = seq(0, fig5_end - fig5_tau, by = fig5_tau)) |>
  dplyr::mutate(amt = amt_mg, evid = 1L, cmt = "central")

fig5_obs <- fig5_subj |>
  tidyr::crossing(time = seq(0, fig5_end, by = 0.25)) |>
  dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")

fig5_events <- dplyr::bind_rows(fig5_dose, fig5_obs) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  dplyr::select(id, time, amt, evid, cmt, WT, CRCL, wt_band, egfr_band)

stopifnot(!anyDuplicated(unique(fig5_events[, c("id", "time", "evid")])))

fig5_sim <- rxode2::rxSolve(
  rxode2::zeroRe(mod),
  events = fig5_events,
  keep   = c("wt_band", "egfr_band")
) |>
  as.data.frame()

fig5_sim |>
  ggplot(aes(time, Cc)) +
  geom_line(linewidth = 0.3) +
  facet_grid(wt_band ~ egfr_band) +
  labs(
    x = "time (h)",
    y = "concentration (mg/L)",
    title = "Figure 5 - ceftazidime 30 mg/kg q12h by weight and renal function",
    caption = "Replicates Figure 5 of Li 2024 (typical-value profiles; panels are eGFR in mL/min/1.73 m^2)."
  ) +
  theme_bw()
```

```{r figure-5-checks}
# rxSolve() returns observation rows only -- there is no `evid` column to
# filter on in the solved output.
fig5_chk <- fig5_sim |>
  dplyr::group_by(wt_band, egfr_band) |>
  dplyr::summarise(
    # Strictly BEFORE tau: the record at t = tau sits after the second dose and
    # would report the accumulated peak, not the first-interval one.
    cmax_first = max(Cc[time < fig5_tau]),
    ctrough_ss = Cc[which.min(abs(time - fig5_end))],
    .groups    = "drop"
  )

# The whole panel grid shares one peak height, because Vd scales with WT^1.
# 30 mg/kg * 70/27.83 = 75.46 mg/L, matching the ~75 mg/L peaks in Figure 5.
expected_peak <- fig5_dose_mgkg * c0_per_mgkg

fig5_chk |>
  tidyr::pivot_wider(names_from = egfr_band, values_from = c(cmax_first, ctrough_ss)) |>
  knitr::kable(
    digits  = 2,
    caption = "Figure 5 first-interval peak and 168 h trough (mg/L) by subgroup."
  )

stopifnot(
  # Deterministic (random effects zeroed), so an exact identity is the right
  # assertion here, not a tolerance band.
  max(abs(fig5_chk$cmax_first - expected_peak)) < 1e-6,
  # Li 2024 Sect. 3.4: within a weight band, higher eGFR lowers concentrations.
  all(
    fig5_chk |>
      dplyr::arrange(wt_band, egfr_band) |>
      dplyr::group_by(wt_band) |>
      dplyr::summarise(mono = all(diff(ctrough_ss) < 0), .groups = "drop") |>
      dplyr::pull(mono)
  ),
  # Li 2024 Sect. 3.4: "ceftazidime concentrations were lower for those weighing
  # less than 10 kg" at a comparable eGFR -- the WT^-0.10 kel scaling.
  all(
    fig5_chk |>
      dplyr::arrange(egfr_band, wt_band) |>
      dplyr::group_by(egfr_band) |>
      dplyr::summarise(mono = all(diff(ctrough_ss) > 0), .groups = "drop") |>
      dplyr::pull(mono)
  )
)
```

Every panel peaks at exactly `r round(expected_peak, 2)` mg/L regardless of
weight, which is the signature of the fixed unit exponent on weight for Vd
(`theta1 = 1`) and matches the roughly 75 mg/L peaks visible across all sixteen
panels of the published figure. Troughs fall monotonically as renal function
improves and rise monotonically with weight, reproducing both directional
claims Li 2024 makes in Section 3.4.

# PKNCA validation

The paper reports no non-compartmental analysis of its own, so PKNCA is used
here to confirm that a single-dose solve of the packaged model reproduces the
closed-form identities of a one-compartment IV bolus (`Cmax = D/Vd`,
`AUCinf = D/CL`, `t1/2 = ln2 * Vd/CL`) and that the resulting half-life sits in
the range the paper quotes for ceftazidime.

```{r pknca}
nca_subj <- tibble::tibble(
  id        = 1:3,
  treatment = c("reference 70 kg, eGFR 116.93",
                "child 15 kg, eGFR 116.93",
                "child 15 kg, eGFR 45"),
  WT        = c(70, 15, 15),
  CRCL      = c(ref_crcl, ref_crcl, 45),
  amt_mg    = c(2000, 750, 750)
)

nca_dose <- nca_subj |>
  dplyr::mutate(time = 0, amt = amt_mg, evid = 1L, cmt = "central")
nca_obs <- nca_subj |>
  tidyr::crossing(time = seq(0, 36, by = 0.05)) |>
  dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")

nca_events <- dplyr::bind_rows(nca_dose, nca_obs) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  dplyr::select(id, time, amt, evid, cmt, WT, CRCL, treatment)

nca_sim <- rxode2::rxSolve(
  rxode2::zeroRe(mod),
  events = nca_events,
  keep   = c("treatment")
) |>
  as.data.frame()

# Only !is.na(Cc) -- adding time > 0 or Cc > 0 would drop the time-zero record
# PKNCA needs to anchor AUC0-*.
sim_nca <- nca_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

stopifnot(nrow(sim_nca) > 0, all(sim_nca$Cc >= 0))

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "ug/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(
  nca_events |> dplyr::filter(evid == 1) |> dplyr::select(id, time, amt, treatment),
  amt ~ time | treatment + id,
  doseu = "mg"
)

intervals <- data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r pknca-compare}
nca_wide <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

# Closed-form expectations for the SAME typical subjects.
expected <- nca_subj |>
  dplyr::mutate(
    vd_i   = vd_ref * (WT / ref_wt),
    cl_i   = cl_ref * (WT / ref_wt)^0.90 * (CRCL / ref_crcl)^0.38,
    cmax_e = amt_mg / vd_i,
    auc_e  = amt_mg / cl_i,
    thalf_e = log(2) * vd_i / cl_i
  ) |>
  dplyr::select(treatment, cmax_e, auc_e, thalf_e)

cmp <- nca_wide |>
  dplyr::left_join(expected, by = "treatment") |>
  dplyr::mutate(
    cmax_pct  = 100 * (cmax - cmax_e) / cmax_e,
    auc_pct   = 100 * (aucinf.obs - auc_e) / auc_e,
    thalf_pct = 100 * (half.life - thalf_e) / thalf_e
  )

cmp |>
  dplyr::select(treatment, cmax, cmax_e, cmax_pct,
                aucinf.obs, auc_e, auc_pct, half.life, thalf_e, thalf_pct) |>
  dplyr::rename(
    "Subject"                 = treatment,
    "Cmax PKNCA (ug/mL)"      = cmax,
    "Cmax D/Vd (ug/mL)"       = cmax_e,
    "Cmax % diff"             = cmax_pct,
    "AUCinf PKNCA (ug*h/mL)"  = aucinf.obs,
    "AUCinf D/CL (ug*h/mL)"   = auc_e,
    "AUCinf % diff"           = auc_pct,
    "t1/2 PKNCA (h)"          = half.life,
    "t1/2 ln2*Vd/CL (h)"      = thalf_e,
    "t1/2 % diff"             = thalf_pct
  ) |>
  knitr::kable(
    digits  = 3,
    caption = "PKNCA output against the closed-form one-compartment IV bolus identities."
  )

# Look rows up BY NAME. PKNCA returns groups in its own (alphabetical) order,
# so positional indexing silently reads the wrong subject.
thalf_of <- function(label) {
  v <- cmp$half.life[cmp$treatment == label]
  if (length(v) != 1L) stop("no unique NCA row for '", label, "'")
  v
}

stopifnot(
  # Both sides use the SAME drawn (here, typical) parameters, so the only
  # difference is numerical: trapezoidal AUC on a 0.05 h grid and a
  # log-linear lambda-z fit. A tight bound is correct and is kept tight.
  max(abs(cmp$cmax_pct))  < 0.01,
  max(abs(cmp$auc_pct))   < 0.10,
  max(abs(cmp$thalf_pct)) < 0.10,
  # Li 2024 Introduction quotes 1.5-2.5 h for normal renal function, extending
  # to 15 h with renal impairment. Both eGFR-116.93 subjects must land in the
  # normal window, and the eGFR-45 subject must be slower than its
  # normal-renal-function twin.
  thalf_of("reference 70 kg, eGFR 116.93") > 1.5,
  thalf_of("reference 70 kg, eGFR 116.93") < 2.5,
  thalf_of("child 15 kg, eGFR 116.93") > 1.5,
  thalf_of("child 15 kg, eGFR 116.93") < 2.5,
  thalf_of("child 15 kg, eGFR 45") > thalf_of("child 15 kg, eGFR 116.93")
)
```

PKNCA reproduces all three closed-form identities to within
`r sprintf("%.3f", max(abs(cmp$cmax_pct)))`% on Cmax,
`r sprintf("%.3f", max(abs(cmp$auc_pct)))`% on AUCinf and
`r sprintf("%.3f", max(abs(cmp$thalf_pct)))`% on half-life, confirming the
packaged model solves to the intended one-compartment structure.

# Target attainment against Table 4

Li 2024 Table 4 is the paper's headline deliverable: the smallest regimen, per
weight and renal-function subgroup, that reaches a probability of target
attainment above 90% for 70% *f*T>MIC, with the free concentration taken as 90%
of total (Li 2024 Sect. 2.7). Three cells are printed as "-" because no regimen
in the paper's search space reached the target. The check below simulates each
recommended regimen to steady state and recomputes PTA.

```{r table4-transcribe}
# Li 2024 Table 4, transcribed verbatim. Doses are mg/kg; tau in hours.
# NA marks the four cells the paper prints as "-" (target unreachable).
tab4 <- tibble::tribble(
  ~wt_band,    ~egfr_band,             ~mic,  ~dose_mgkg, ~tau,
  "<10 kg",    "30-60 (moderate)",        2,        17.5,   12,
  "<10 kg",    "30-60 (moderate)",        4,        35.0,   12,
  "<10 kg",    "30-60 (moderate)",        8,        22.5,    8,
  "<10 kg",    "30-60 (moderate)",       16,        25.0,    6,
  "<10 kg",    "60-90 (mild)",            2,        32.5,   12,
  "<10 kg",    "60-90 (mild)",            4,        17.5,    8,
  "<10 kg",    "60-90 (mild)",            8,        32.5,    8,
  "<10 kg",    "60-90 (mild)",           16,        35.0,    6,
  "<10 kg",    "90-120 (normal)",         2,        50.0,   12,
  "<10 kg",    "90-120 (normal)",         4,        25.0,    8,
  "<10 kg",    "90-120 (normal)",         8,        22.5,    6,
  "<10 kg",    "90-120 (normal)",        16,          NA,   NA,
  "<10 kg",    "120-200 (augmented)",     2,        20.0,    8,
  "<10 kg",    "120-200 (augmented)",     4,        16.0,    6,
  "<10 kg",    "120-200 (augmented)",     8,        30.0,    6,
  "<10 kg",    "120-200 (augmented)",    16,          NA,   NA,
  "10-30 kg",  "30-60 (moderate)",        2,        12.5,   12,
  "10-30 kg",  "30-60 (moderate)",        4,        22.5,   12,
  "10-30 kg",  "30-60 (moderate)",        8,        50.0,   12,
  "10-30 kg",  "30-60 (moderate)",       16,        32.5,    8,
  "10-30 kg",  "60-90 (mild)",            2,        20.0,   12,
  "10-30 kg",  "60-90 (mild)",            4,        40.0,   12,
  "10-30 kg",  "60-90 (mild)",            8,        25.0,    8,
  "10-30 kg",  "60-90 (mild)",           16,        27.5,    6,
  "10-30 kg",  "90-120 (normal)",         2,        30.0,   12,
  "10-30 kg",  "90-120 (normal)",         4,        17.5,    8,
  "10-30 kg",  "90-120 (normal)",         8,        32.5,    8,
  "10-30 kg",  "90-120 (normal)",        16,        32.5,    6,
  "10-30 kg",  "120-200 (augmented)",     2,        12.5,    8,
  "10-30 kg",  "120-200 (augmented)",     4,        25.0,    8,
  "10-30 kg",  "120-200 (augmented)",     8,        22.5,    6,
  "10-30 kg",  "120-200 (augmented)",    16,          NA,   NA,
  "30-50 kg",  "30-60 (moderate)",        2,        10.0,   12,
  "30-50 kg",  "30-60 (moderate)",        4,        20.0,   12,
  "30-50 kg",  "30-60 (moderate)",        8,        40.0,   12,
  "30-50 kg",  "30-60 (moderate)",       16,        30.0,    8,
  "30-50 kg",  "60-90 (mild)",            2,        17.5,   12,
  "30-50 kg",  "60-90 (mild)",            4,        35.0,   12,
  "30-50 kg",  "60-90 (mild)",            8,        22.5,    8,
  "30-50 kg",  "60-90 (mild)",           16,        25.0,    6,
  "30-50 kg",  "90-120 (normal)",         2,        25.0,   12,
  "30-50 kg",  "90-120 (normal)",         4,        47.5,   12,
  "30-50 kg",  "90-120 (normal)",         8,        30.0,    8,
  "30-50 kg",  "90-120 (normal)",        16,        30.0,    6,
  "30-50 kg",  "120-200 (augmented)",     2,        42.5,   12,
  "30-50 kg",  "120-200 (augmented)",     4,        20.0,    8,
  "30-50 kg",  "120-200 (augmented)",     8,        20.0,    6,
  "30-50 kg",  "120-200 (augmented)",    16,        40.0,    6,
  "50-70 kg",  "30-60 (moderate)",        2,        10.0,   12,
  "50-70 kg",  "30-60 (moderate)",        4,        17.5,   12,
  "50-70 kg",  "30-60 (moderate)",        8,        35.0,   12,
  "50-70 kg",  "30-60 (moderate)",       16,        27.5,    8,
  "50-70 kg",  "60-90 (mild)",            2,        15.0,   12,
  "50-70 kg",  "60-90 (mild)",            4,        30.0,   12,
  "50-70 kg",  "60-90 (mild)",            8,        20.0,    8,
  "50-70 kg",  "60-90 (mild)",           16,        22.5,    6,
  "50-70 kg",  "90-120 (normal)",         2,        22.5,   12,
  "50-70 kg",  "90-120 (normal)",         4,        45.0,   12,
  "50-70 kg",  "90-120 (normal)",         8,        25.0,    8,
  "50-70 kg",  "90-120 (normal)",        16,        27.5,    6,
  "50-70 kg",  "120-200 (augmented)",     2,        40.0,   12,
  "50-70 kg",  "120-200 (augmented)",     4,        20.0,    8,
  "50-70 kg",  "120-200 (augmented)",     8,        20.0,    6,
  "50-70 kg",  "120-200 (augmented)",    16,        37.5,    6
)

# Three cells are printed as "-": <10 kg at eGFR 90-120 and 120-200, and
# 10-30 kg at eGFR 120-200, all at MIC 16 mg/L. The Li 2024 Discussion states
# exactly this set ("under 10 kg with an eGFR of 90-200 ... or those weighing
# 10-30 kg with an eGFR of 120-200"), which corroborates the transcription.
stopifnot(nrow(tab4) == 64L, sum(is.na(tab4$dose_mgkg)) == 3L)

tab4_gated <- tab4 |>
  dplyr::filter(!is.na(dose_mgkg)) |>
  dplyr::mutate(cell = dplyr::row_number())
stopifnot(nrow(tab4_gated) == 61L)
```

```{r table4-simulate}
pd_target  <- 0.70   # Li 2024 Sect. 2.7: 70% fT>MIC
free_frac  <- 0.90   # Li 2024 Sect. 2.7: free concentration is 90% of total
n_per_cell <- 200L   # per-arm cohort; the 200/arm cap applies per cell
ss_start   <- 72     # dose for 72 h before the evaluated interval (>= 9 half-lives)

# Covariate values are sampled UNIFORMLY within each subgroup band. Li 2024
# does not state how it drew covariates inside a band for the Monte Carlo, so
# this is an assumption (recorded in the Errata).
set.seed(4711)
cells <- tab4_gated |>
  dplyr::left_join(wt_bands,   by = "wt_band") |>
  dplyr::left_join(egfr_bands, by = "egfr_band")

subj <- cells |>
  tidyr::crossing(rep = seq_len(n_per_cell)) |>
  dplyr::mutate(
    id   = dplyr::row_number(),
    WT   = runif(dplyr::n(), wt_lo, wt_hi),
    CRCL = runif(dplyr::n(), egfr_lo, egfr_hi),
    amt  = dose_mgkg * WT
  )

dose_rows <- subj |>
  dplyr::group_by(id) |>
  dplyr::reframe(
    time = seq(0, ss_start, by = dplyr::first(tau)),
    amt  = dplyr::first(amt), evid = 1L, cmt = "central"
  )

# Observations span the final (steady-state) dosing interval only.
obs_rows <- subj |>
  dplyr::group_by(id) |>
  dplyr::reframe(
    time = ss_start + seq(0, dplyr::first(tau), by = 0.25),
    amt  = NA_real_, evid = 0L, cmt = "central"
  )

pta_events <- dplyr::bind_rows(dose_rows, obs_rows) |>
  dplyr::left_join(subj |> dplyr::select(id, WT, CRCL, cell), by = "id") |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  dplyr::select(id, time, amt, evid, cmt, WT, CRCL, cell)

stopifnot(!anyDuplicated(unique(pta_events[, c("id", "time", "evid")])))

pta_sim <- rxode2::rxSolve(mod, events = pta_events, keep = c("cell")) |>
  as.data.frame()
```

`fT>MIC` is computed two ways. The closed form uses each simulated subject's
own `cl` and `vc` (returned as columns by `rxSolve`) and is exact for a
one-compartment IV bolus at steady state; the grid form reads the crossing time
off the simulated profile by log-linear interpolation. Agreement between them
confirms the closed form is measuring the solved model, not an idealisation of
it.

```{r table4-ftmic}
# Per-subject parameters, taken from the solved output.
pars <- pta_sim |>
  dplyr::group_by(id) |>
  dplyr::summarise(cl = dplyr::first(cl), vc = dplyr::first(vc), .groups = "drop") |>
  dplyr::left_join(subj |> dplyr::select(id, cell, amt, tau, mic), by = "id") |>
  dplyr::mutate(
    kel     = cl / vc,
    cmax_ss = (amt / vc) / (1 - exp(-kel * tau)),
    # Time above MIC within one steady-state interval, capped at tau.
    t_above = pmin(tau, pmax(0, log(free_frac * cmax_ss / mic) / kel)),
    ft_closed = t_above / tau
  )

# Grid form: the last observation at which the free concentration exceeds MIC,
# refined by log-linear interpolation onto the next grid point. The profile
# within one steady-state interval starts at the dose and decays monotonically,
# so the points above MIC are a leading run and `nab` (their count) is also the
# index of the last one -- which lets every step below be a scalar summary
# rather than a per-group R closure over 6100 groups.
grid_ft <- pta_sim |>
  dplyr::left_join(subj |> dplyr::select(id, tau, mic), by = "id") |>
  dplyr::mutate(t_rel = time - ss_start, cfree = free_frac * Cc) |>
  dplyr::arrange(id, t_rel) |>
  dplyr::group_by(id) |>
  dplyr::summarise(
    tau  = dplyr::first(tau),
    mic  = dplyr::first(mic),
    ntot = dplyr::n(),
    nab  = sum(cfree > mic),
    # Index-clamped neighbours bracketing the crossing. When nab is 0 or ntot
    # these are unused (case_when below returns the endpoint) but must still be
    # in range.
    c_lo = cfree[pmin(pmax(nab, 1L), ntot)],
    t_lo = t_rel[pmin(pmax(nab, 1L), ntot)],
    c_hi = cfree[pmin(nab + 1L, ntot)],
    t_hi = t_rel[pmin(nab + 1L, ntot)],
    .groups = "drop"
  ) |>
  dplyr::mutate(
    t_grid = dplyr::case_when(
      nab == 0L   ~ 0,
      nab == ntot ~ tau,
      # Log-linear interpolation is exact for a mono-exponential decay.
      TRUE ~ t_lo + (t_hi - t_lo) * (log(c_lo) - log(mic)) / (log(c_lo) - log(c_hi))
    ),
    ft_grid = t_grid / tau
  )

ft <- pars |> dplyr::left_join(grid_ft |> dplyr::select(id, ft_grid), by = "id")

stopifnot(
  nrow(ft) == nrow(subj),
  !anyNA(ft$ft_closed), !anyNA(ft$ft_grid),
  # Both sides describe the SAME solved subject; the only difference is grid
  # resolution, so this bound is numerical and stays tight.
  max(abs(ft$ft_closed - ft$ft_grid)) < 0.005
)
```

```{r table4-pta}
pta <- ft |>
  dplyr::group_by(cell) |>
  dplyr::summarise(
    pta       = 100 * mean(ft_closed >= pd_target),
    ft_median = median(ft_closed),
    .groups   = "drop"
  ) |>
  dplyr::left_join(tab4_gated, by = "cell")

pta_wide <- pta |>
  dplyr::mutate(
    label = sprintf("%s mg/kg q%gh (%.0f%%)", format(dose_mgkg, trim = TRUE), tau, pta),
    mic_col = paste0("MIC ", mic, " mg/L")
  ) |>
  dplyr::select(wt_band, egfr_band, mic_col, label) |>
  tidyr::pivot_wider(names_from = mic_col, values_from = label) |>
  dplyr::arrange(wt_band, egfr_band)

pta_wide |>
  dplyr::rename("Weight" = wt_band, "eGFR (mL/min/1.73 m^2)" = egfr_band) |>
  knitr::kable(
    caption = paste(
      "Li 2024 Table 4 recommended regimens, each followed by the PTA this",
      "model reproduces for 70% fT>MIC (free = 90% of total). The paper's",
      "acceptance criterion was PTA > 90%; see the known deviation below."
    )
  )

data.frame(
  statistic = c("cells checked", "minimum PTA (%)", "median PTA (%)",
                "maximum PTA (%)", "cells with PTA >= 90%",
                "lowest per-cell median fT>MIC"),
  value = c(nrow(pta), round(min(pta$pta), 1), round(median(pta$pta), 1),
            round(max(pta$pta), 1), sum(pta$pta >= 90),
            round(min(pta$ft_median), 3))
) |>
  knitr::kable(caption = "Summary of the Table 4 target-attainment reproduction.")
```

```{r table4-gate}
# KNOWN DEVIATION -- see the narrative below. The model reproduces the SHAPE of
# Table 4 but lands PTA a few points BELOW the paper's own > 90% criterion
# (measured median 87.5%, range 77-94.5% at n = 200/cell; 87 / 76 / 94.5 at
# n = 100/cell). The gate is therefore written around what IS reproduced, and
# the shortfall is reported rather than tuned away.
#
# The bounds are two-sided on purpose, and both arms were checked to go red:
#   * lower arm  -- exposure-deflating errors. (A mis-transcribed CL or Vd is
#     in fact caught earlier and harder, by the PKNCA identity gate and the
#     Figure 5 exact-peak gate respectively, both of which fail outright.)
#   * upper arm  -- reading Table 3's omega^2 rows as standard deviations
#     instead of variances shrinks IIV about 25-fold and drives PTA to a
#     median of 100.0% (min 99.0), which this bound rejects.
stopifnot(
  # The median subject of every recommended regimen attains the target. This is
  # the structural claim of Table 4 and it holds in all 61 gated cells.
  min(pta$ft_median) >= pd_target,
  # Cohort PTA is high everywhere, with headroom below the observed minimum.
  min(pta$pta) >= 65,
  # Two-sided band on the central tendency; 97 is the arm that catches an
  # IIV misread, 75 the arm that catches exposure deflation.
  median(pta$pta) >= 75, median(pta$pta) <= 97
)
```

## Known deviation: PTA lands a few points below the paper's criterion

The model reproduces the *structure* of Table 4 but not its headline threshold.
Every one of the 61 recommended regimens attains 70% *f*T>MIC for the median
subject, and the internal logic of the table is reproduced exactly -- doubling
the MIC at a fixed interval doubles the recommended dose (`<10 kg` / eGFR 30-60
goes 17.5 to 35 mg/kg q12h; `30-50 kg` / eGFR 30-60 goes 10 to 20 to 40 mg/kg
q12h), the dose rises monotonically with eGFR within a weight band, and the
regimen shortens from q12h to q8h to q6h as the MIC climbs. What does *not*
reproduce is the acceptance threshold: cohort PTA comes out at a median of
about 87% with a range of roughly 77-95%, where the paper's stated criterion
for including a regimen in Table 4 is PTA > 90%.

The shortfall is uniform across MIC, dosing interval and subgroup, which argues
for a single systematic difference rather than a transcription error in any one
cell. Three candidate mechanisms, none of which the paper reports in enough
detail to settle:

* **How covariates were drawn inside each subgroup.** Replacing the uniform
  within-band draw used here with the band midpoint changes nothing material
  (median 87.1% against 87.2%), so this is *not* the driver -- but a draw from
  the observed 88-subject cohort restricted to each band, which is what a
  resampling-style Monte Carlo would do, would give different tail behaviour.
* **Infusion duration.** Li 2024 never reports one. A finite infusion lowers the
  peak, which would push PTA further down, so this cannot explain a shortfall.
* **The statistic is intrinsically knife-edged at the optimum.** Because Table 4
  reports the *smallest* regimen meeting the criterion, the 10th percentile of
  the *f*T>MIC distribution sits essentially on the target by construction. A
  few percent difference in the assumed variability therefore moves PTA by
  several points in either direction, and PTA near an optimizer's boundary is
  the least robust quantity a paper of this kind publishes.

Rather than widen the gate until the disagreement disappears, the check above
asserts what is robustly reproduced (median-subject attainment in every cell,
plus a two-sided band on cohort PTA) and leaves the measured PTA visible in the
table so a reader can see the gap.

The three cells Li 2024 prints as "-" are excluded from the gate: the paper
states that no regimen in its search space reached PTA > 90% there, but it does
not report the upper bound of that search grid, so there is no well-defined
regimen to simulate against. All three sit at MIC 16 mg/L in the smaller-weight,
higher-eGFR corner of the grid, which is where the `kel` scaling
(`WT^-0.10 * eGFR^0.38`) makes elimination fastest -- the same corner the
Discussion singles out as requiring an elevated dose.

# Assumptions and deviations

- **Infusion duration is not reported.** Li 2024 Sect. 2.2 states only that
  ceftazidime was "administered intravenously"; no infusion duration appears
  anywhere in the paper. Doses are therefore simulated as IV boluses into
  `central`. The published Figure 5 supports this: its peaks sit at
  approximately 75 mg/L for a 30 mg/kg dose, which is exactly
  `30 * 70 / 27.83 = 75.46` mg/L, the instantaneous-input value. A finite
  infusion would lower the peak and slightly lengthen *f*T>MIC, so the bolus
  assumption is the conservative one for target attainment.
- **Covariate sampling within a subgroup band.** Li 2024 does not state how
  weight and eGFR were drawn inside each subgroup for the Monte Carlo
  simulations. Both are sampled uniformly across the band here. The `<10 kg`
  band is truncated below at 2.80 kg, the cohort minimum from Table 1, rather
  than at zero. Substituting the band midpoint for the uniform draw was checked
  and moves the median reproduced PTA by 0.1 percentage points, so the choice
  is not load-bearing.
- **Target attainment is a documented deviation, not a match.** The reproduced
  PTA for the Table 4 regimens has a median of about 87% against the paper's
  stated PTA > 90% acceptance criterion. See "Known deviation" above for the
  measured values and the candidate mechanisms; no parameter was adjusted to
  close the gap.
- **Reference weight is an extrapolation.** The 70 kg normalising weight in
  Eqs. 5 and 6 is a nominal adult reference, not a cohort value -- the cohort
  median is 18.35 kg and only one subject reaches 95 kg. The reported 27.83 L
  and 7.76 L/h therefore describe a hypothetical 70 kg subject rather than an
  observed one. This is the paper's own parameterisation and is reproduced
  as published.
- **Residual error is additive only.** Li 2024 Sect. 3.2 reports that the
  additive model outperformed the proportional and combined models, so the
  packaged model carries `addSd` and no proportional term. With an additive SD
  of 1.20 mg/L against a median observed concentration of 3.35 mg/L, the
  relative error at low concentrations is large; that is what the source
  reports.
- **IIV values are variances, not CV%.** Li 2024 Table 3 labels the rows
  `omega^2_Vd` and `omega^2_CL` while the footnote defines the unsquared
  `omega` as the "square root of inter-individual variance". The tabulated
  0.04 and 0.06 are therefore log-scale variances and are used as printed
  (equivalent to roughly 20% and 25% CV); no CV-to-variance back-transformation
  is applied.
- **Serum creatinine unit trap.** Table 1 reports serum creatinine in umol/L
  (median 33.9) but the modified Schwartz equation consumes mg/dL. The median
  subject reproduces the reported median eGFR only under the mg/dL reading
  (`0.413 * 111 / (33.9 / 88.4) = 119.6` against a reported median of 116.93),
  which settles the input unit. Users supplying `CRCL` to this model must use
  mL/min/1.73 m^2 computed that way.

# Errata and source gaps

- **Supplementary Tables S1 and S2 are not available.** Li 2024 cites
  Supplementary Table S1 (per-covariate hypothesis-test results) and
  Supplementary Table S2 (the full PTA grid across all evaluated regimens). The
  Frontiers supplementary bundle is not part of the PMC open-access deposit
  (`PMC11631598/supplementaryFiles` returns only the article figures and the
  main-text Table 1), and the publisher's supplementary endpoints did not
  resolve. Neither table contains model parameters: every value in the model
  file comes from main-text Table 3 and Equations 5-6, and Table 4 supplies the
  recommended-regimen grid used for validation above. The covariates listed in
  `covariatesDataExcluded` are therefore documented from the Methods candidate
  list and the Section 3.2 narrative rather than from S1's statistics.
- **Equations 5 and 6 are vector-typeset.** Both are stacked fractions in the
  published PDF and survive neither the markdown trim nor a plain text
  extraction; `pdftotext -layout` recovers them intact and was the source used.
- **A dose off the search grid.** Table 4 uses a 2.5 mg/kg dose grid with a
  10 mg/kg floor in every cell except `<10 kg` / eGFR 120-200 / MIC 4 mg/L,
  which prints 16 mg/kg q6h. That value is off the grid every other cell
  follows and may be a typographical error for 15 or 17.5, but it is
  reproduced as printed and is not adjusted here.
- **Bootstrap CI typographical error.** Li 2024 Table 3 gives the bootstrap 95%
  CI for theta2 as "0.81-10.03", which repeats the upper bound of the CL row
  and is implausible for an exponent whose point estimate is 0.90 with a 6.41%
  RSE. The point estimate, which is what the model uses, is unaffected.
