---
title: "Paroxetine (Li 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Paroxetine (Li 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
suppressPackageStartupMessages({
  library(nlmixr2lib)
  library(PKNCA)
  library(rxode2)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Li_2022_paroxetine"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Li_2022_paroxetine"))$description`
* Article (open access): <https://doi.org/10.3389/fphar.2022.966622>

# Population

Li 2022 analysed retrospective therapeutic-drug-monitoring (TDM) data from
psychiatric inpatients treated with paroxetine at the Affiliated Brain Hospital
of Guangzhou Medical University, China, between January 2019 and May 2021. The
analysis set held 184 patients (103 female, 56%; 81 male) and 372 serum
paroxetine concentrations (median 60.55 ng/mL, range 2.62-447.87). Median age
was 37.5 years (range 15-90), median body weight 60 kg (printed range 58-96),
and median height 164.5 cm (Li 2022 Table 1). Daily doses ranged from 20 to
75 mg and were adjusted during treatment; 85.5% of the concentrations were on
immediate-release tablets and 14.5% on 25 mg sustained-release tablets.
Samples were drawn 10-22.5 h after the most recent dose, so the data are
essentially troughs. Co-medications included olanzapine (16.8%),
tandospirone (14.1%), risperidone (13.0%) and metoprolol (5.4%); none was
retained as a covariate, and CYP2D6 genotype was not retained either.

The same information is available programmatically via
`readModelDb("Li_2022_paroxetine")()$population`.

# Source trace

Li 2022 gives the final model as four printed equations (Section 3.2.1):

* `CL/F (L/h) = 21.2 * (dose/40)^-1.03`
* `V/F (L) = 8850 * (1 - 0.666 * formulation)`
* `Ka (1/h) = 0.908`
* `F1 = 1 + (0.475 * SEX)`

with `formulation = 1` for the immediate-release tablet and `SEX = 1` for
female (Methods 2.3.2).

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` | `fixed(log(0.908))` | Table 2 'Ka (1/h) 0.908 fixed'; Methods 2.3.1 (literature value) |
| `lcl` | `log(21.2)` | Table 2 'CL/F (L/h) 21.2' (RSE 7.2%) |
| `lvc` | `log(8850)` | Table 2 'V/F (L) 8,850' (RSE 17.2%) |
| `e_dose_cl` | `-1.03` | Table 2 'theta CL-dosage' (RSE 5.7%); Section 3.2.1 CL/F equation, reference 40 mg/day |
| `e_form_ir_vc` | `0.666` | Table 2 'theta V-formulation' (RSE 9.7%); Section 3.2.1 V/F equation |
| `e_sexf_fdepot` | `0.475` | Table 2 'theta F-sex' (RSE 25.3%); Section 3.2.1 F1 equation |
| `etalcl` | `0.2385136` | Table 2 'IIV (CV%)' CL/F 51.9, as `log(1 + 0.519^2)` |
| `etalvc` | `0.5289946` | Table 2 'IIV (CV%)' V/F 83.5, as `log(1 + 0.835^2)` |
| `propSd` | `0.3047950` | Table 2 'PRO (CV%) 0.0929', read as a variance, `sqrt(0.0929)` |
| `cl = exp(lcl + etalcl) * (DOSE/40)^e_dose_cl` | n/a | Section 3.2.1; Eq. 1 (exponential IIV) and Eq. 3 (power covariate) |
| `vc = exp(lvc + etalvc) * (1 - e_form_ir_vc * FORM_PAROXETINE_IR)` | n/a | Section 3.2.1; Eq. 4 (linear categorical covariate) |
| `f(depot) = 1 + e_sexf_fdepot * SEXF` | n/a | Section 3.2.1; Eq. 6 |
| One compartment, first-order absorption | n/a | Methods 2.3.1 (NONMEM ADVAN2) |
| `Cc ~ prop(propSd)` | n/a | Table 2 'PRO, proportional residual error' |

```{r model}
mod <- readModelDb("Li_2022_paroxetine")
mod_typical <- mod |> rxode2::zeroRe()
```

# Typical-value replication of Figure 3 (dose and sex)

Figure 3A of Li 2022 shows the typical trough concentration of the
immediate-release tablet after 20-60 mg once daily in males and females, and
Figure 3B the same values divided by the daily dose. The Methods describe the
simulation as 100 days of once-daily dosing. The figure is reproduced instead
by the trough 24 h after the 30th dose (720 h), as the comparison below shows,
so both time points are simulated.

```{r fig3-events}
doses <- c(20, 30, 40, 50, 60)
typ_cohort <- tidyr::expand_grid(
  DOSE = doses,
  SEXF = c(0, 1),
  FORM_PAROXETINE_IR = c(1, 0)
) |>
  dplyr::mutate(id = dplyr::row_number())

# 100 days of once-daily dosing into the depot; troughs are read at 720 h
# (24 h after the 30th dose) and 2400 h (24 h after the 100th dose). Dosing
# is into the depot, so Cc is continuous at the dose times and a sample that
# coincides with a dose is still the pre-dose trough.
make_events <- function(cohort, obs_times, n_doses = 100) {
  dose_rows <- cohort |>
    dplyr::mutate(
      time = 0,
      evid = 1L,
      amt = DOSE,
      cmt = "depot",
      ii = 24,
      addl = n_doses - 1L
    )
  obs_rows <- cohort |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(evid = 0L, amt = NA_real_, cmt = "central", ii = 0, addl = 0L)
  # The event columns lead: with a covariate (here DOSE) as the first column
  # of the table, rxSolve() does not pick it up and stops with "required
  # parameter(s): DOSE".
  dplyr::bind_rows(dose_rows, obs_rows) |>
    dplyr::select(id, time, evid, amt, cmt, ii, addl, dplyr::everything()) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

ev_typ <- make_events(typ_cohort, obs_times = c(720, 2400))
sim_typ <- rxode2::rxSolve(
  mod_typical,
  events = ev_typ,
  keep = c("DOSE", "SEXF", "FORM_PAROXETINE_IR")
) |>
  as.data.frame()
```

```{r fig3-compare}
# Figure 3A digitised by the maintainers (immediate-release tablet). Reading
# uncertainty is about 1 ng/mL (one pixel at 300 dpi = 0.46 ng/mL).
fig3a <- tibble::tribble(
  ~DOSE, ~SEXF, ~published,
  20, 0, 16.4,
  30, 0, 39.6,
  40, 0, 72.4,
  50, 0, 114.8,
  60, 0, 164.4,
  20, 1, 24.6,
  30, 1, 59.7,
  40, 1, 107.9,
  50, 1, 169.9,
  60, 1, 243.2
)

fig3_cmp <- sim_typ |>
  dplyr::filter(FORM_PAROXETINE_IR == 1) |>
  dplyr::select(DOSE, SEXF, time, Cc) |>
  tidyr::pivot_wider(names_from = time, values_from = Cc, names_prefix = "t") |>
  dplyr::inner_join(fig3a, by = c("DOSE", "SEXF")) |>
  dplyr::mutate(
    sex = ifelse(SEXF == 1, "female", "male"),
    pct_day30 = 100 * (t720 / published - 1),
    pct_day100 = 100 * (t2400 / published - 1),
    cd_ratio_day30 = t720 / DOSE
  )
stopifnot(nrow(fig3_cmp) == nrow(fig3a))

fig3_cmp |>
  dplyr::select(sex, DOSE, published, t720, pct_day30, t2400, pct_day100, cd_ratio_day30) |>
  dplyr::rename(
    "Sex" = sex,
    "Daily dose (mg)" = DOSE,
    "Figure 3A (ng/mL)" = published,
    "Model, 720 h (ng/mL)" = t720,
    "Diff. 720 h (%)" = pct_day30,
    "Model, 2400 h (ng/mL)" = t2400,
    "Diff. 2400 h (%)" = pct_day100,
    "C/D at 720 h" = cd_ratio_day30
  ) |>
  knitr::kable(digits = 2, caption = "Typical immediate-release trough concentrations against Figure 3A of Li 2022.")
```

The trough at 720 h matches every digitised point within
`r sprintf("%.1f", max(abs(fig3_cmp$pct_day30)))`%. At 2400 h the high-dose
points are 3-4% above the figure, because the larger doses have the lowest
clearance and are still accumulating at day 30. The 1.475-fold female/male
ratio is the sex effect on relative bioavailability, and the C/D column
reproduces Figure 3B (male 0.84 to 2.75, female 1.24 to 4.06 as digitised).

```{r fig3-gate}
# Deterministic typical-value check: no random effects are involved, so the
# only error is the digitisation of Figure 3A (about 1 ng/mL, which is 6% at
# the 16 ng/mL point). A wrong exponent, centring dose, V/F coding or sex
# coding moves these points by 10% or more.
stopifnot(
  max(abs(fig3_cmp$pct_day30)) < 5,
  abs(median(fig3_cmp$pct_day30)) < 2
)
```

```{r fig3-plot}
ggplot(fig3_cmp, aes(DOSE, t720, colour = sex)) +
  geom_line() +
  geom_point() +
  geom_point(aes(y = published), shape = 4, size = 3) +
  geom_hline(yintercept = c(20, 65), linetype = "dotted") +
  geom_hline(yintercept = 120, colour = "orange") +
  labs(
    x = "Daily dose (mg/day)",
    y = "Trough concentration (ng/mL)",
    colour = NULL,
    title = "Figure 3A - typical immediate-release troughs",
    caption = "Lines: model at 720 h. Crosses: digitised Figure 3A of Li 2022."
  )
```

# Typical-value replication of Figure 5 (formulation)

Figure 5 of Li 2022 compares the typical immediate-release and
sustained-release profiles after 50 mg once daily. The paper's Methods code
the immediate-release tablet as 1, which gives it the smaller V/F
(2956 L against 8850 L). The figure confirms that coding: the
immediate-release curve swings by about 13 ng/mL across the dosing interval,
the sustained-release curve by about 6 ng/mL.

```{r fig5}
fig5_cohort <- typ_cohort |>
  dplyr::filter(DOSE == 50)
ev_fig5 <- make_events(fig5_cohort, obs_times = seq(696, 744, by = 0.5))
sim_fig5 <- rxode2::rxSolve(
  mod_typical,
  events = ev_fig5,
  keep = c("SEXF", "FORM_PAROXETINE_IR")
) |>
  as.data.frame() |>
  dplyr::mutate(
    sex = ifelse(SEXF == 1, "Female", "Male"),
    formulation = ifelse(FORM_PAROXETINE_IR == 1, "IR-T", "SR-T")
  )

ggplot(sim_fig5, aes(time - 696, Cc, linetype = formulation)) +
  geom_line() +
  facet_wrap(~sex) +
  labs(
    x = "Time after the 30th dose (h)",
    y = "Paroxetine (ng/mL)",
    linetype = NULL,
    title = "Figure 5 - 50 mg once daily, days 30-31",
    caption = "Replicates Figure 5 of Li 2022 (typical values)."
  )

fig5_summary <- sim_fig5 |>
  dplyr::filter(time >= 720, time <= 744) |>
  dplyr::group_by(sex, formulation) |>
  dplyr::summarise(
    trough = Cc[time == 720],
    peak = max(Cc),
    swing = peak - trough,
    .groups = "drop"
  )
knitr::kable(fig5_summary, digits = 1, caption = "Typical trough, peak and peak-trough swing over the 24 h after the 30th dose.")
```

```{r fig5-gate}
swing <- function(s, f) fig5_summary$swing[fig5_summary$sex == s & fig5_summary$formulation == f]
trough <- function(s, f) fig5_summary$trough[fig5_summary$sex == s & fig5_summary$formulation == f]
stopifnot(nrow(fig5_summary) == 4L)
# Figure 5 (digitised): male IR-T trough ~115 and swing ~13 ng/mL, SR-T
# trough ~89 and swing ~6 ng/mL. The swing is roughly F * dose / (V/F), so it
# is the direct test of which formulation carries the 0.334 factor on V/F.
stopifnot(
  abs(trough("Male", "IR-T") / 115 - 1) < 0.05,
  abs(trough("Male", "SR-T") / 89 - 1) < 0.05,
  swing("Male", "IR-T") / swing("Male", "SR-T") > 2
)
```

At true steady state the two tablets have the same average concentration,
because formulation enters only V/F. The lower sustained-release trough in
Figures 4 and 5 of the paper comes from the larger V/F: at 50 mg the
sustained-release half-life is `r sprintf("%.0f", log(2) * 8850 / (21.2 * (50 / 40)^-1.03) / 24)` days, so at day 30 it is still
accumulating. By day 100 the typical sustained-release trough
(`r sprintf("%.0f", sim_typ$Cc[sim_typ$DOSE == 50 & sim_typ$SEXF == 0 & sim_typ$FORM_PAROXETINE_IR == 0 & sim_typ$time == 2400])` ng/mL for a male)
is slightly above the immediate-release one
(`r sprintf("%.0f", sim_typ$Cc[sim_typ$DOSE == 50 & sim_typ$SEXF == 0 & sim_typ$FORM_PAROXETINE_IR == 1 & sim_typ$time == 2400])` ng/mL).

# Stochastic replication of Figure 4

Figure 4 of Li 2022 shows box plots of simulated troughs (individual
predictions, no residual error) for 25 and 50 mg once daily by sex and
formulation. A virtual cohort of 100 patients per arm (8 arms) is simulated
with between-subject variability and the trough read at 720 h, matching the
time point that reproduces Figures 3 and 5.

```{r fig4}
rxode2::rxSetSeed(20220912)
n_per_arm <- 100L
fig4_cohort <- tidyr::expand_grid(
  DOSE = c(25, 50),
  SEXF = c(0, 1),
  FORM_PAROXETINE_IR = c(1, 0),
  rep = seq_len(n_per_arm)
) |>
  dplyr::mutate(id = dplyr::row_number()) |>
  dplyr::select(-rep)
ev_fig4 <- make_events(fig4_cohort, obs_times = 720)
stopifnot(!anyDuplicated(unique(ev_fig4[, c("id", "time", "evid")])))

sim_fig4 <- rxode2::rxSolve(
  mod,
  events = ev_fig4,
  keep = c("DOSE", "SEXF", "FORM_PAROXETINE_IR")
) |>
  as.data.frame() |>
  dplyr::mutate(
    sex = ifelse(SEXF == 1, "female", "male"),
    formulation = ifelse(FORM_PAROXETINE_IR == 1, "immediate-release tablet", "sustained-release tablet"),
    dose = paste(DOSE, "mg")
  )
stopifnot(!anyNA(sim_fig4$Cc))

ggplot(sim_fig4, aes(sex, Cc, fill = formulation)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~dose) +
  geom_hline(yintercept = c(20, 65), linetype = "dashed") +
  geom_hline(yintercept = 120, colour = "red", linetype = "dashed") +
  labs(
    x = NULL,
    y = "Simulated trough (ng/mL)",
    fill = NULL,
    title = "Figure 4 - simulated troughs at 720 h",
    caption = "Replicates Figure 4 of Li 2022 (100 virtual patients per arm, individual predictions)."
  ) +
  theme(legend.position = "bottom")
```

```{r fig4-compare}
# Figure 4 box-plot medians digitised by the maintainers.
fig4_pub <- tibble::tribble(
  ~DOSE, ~SEXF, ~FORM_PAROXETINE_IR, ~published_median,
  25, 0, 1, 23.8,
  25, 0, 0, 21.0,
  25, 1, 1, 36.5,
  25, 1, 0, 33.0,
  50, 0, 1, 98.9,
  50, 0, 0, 75.0,
  50, 1, 1, 147.3,
  50, 1, 0, 110.8
)
fig4_cmp <- sim_fig4 |>
  dplyr::group_by(DOSE, SEXF, FORM_PAROXETINE_IR, sex, formulation) |>
  dplyr::summarise(sim_median = median(Cc), .groups = "drop") |>
  dplyr::inner_join(fig4_pub, by = c("DOSE", "SEXF", "FORM_PAROXETINE_IR")) |>
  dplyr::mutate(pct_diff = 100 * (sim_median / published_median - 1))
stopifnot(nrow(fig4_cmp) == 8L)

fig4_cmp |>
  dplyr::select(DOSE, sex, formulation, published_median, sim_median, pct_diff) |>
  dplyr::rename(
    "Daily dose (mg)" = DOSE,
    "Sex" = sex,
    "Formulation" = formulation,
    "Figure 4 median (ng/mL)" = published_median,
    "Simulated median (ng/mL)" = sim_median,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = 1, caption = "Median simulated trough at 720 h against Figure 4 of Li 2022.")

ratio_ir_sr <- sim_fig4 |>
  dplyr::filter(DOSE == 50) |>
  dplyr::group_by(FORM_PAROXETINE_IR) |>
  dplyr::summarise(m = median(Cc / (1 + 0.475 * SEXF)), .groups = "drop")
ratio_ir_sr <- ratio_ir_sr$m[ratio_ir_sr$FORM_PAROXETINE_IR == 1] /
  ratio_ir_sr$m[ratio_ir_sr$FORM_PAROXETINE_IR == 0]
```

Across the eight arms the simulated medians differ from the digitised
Figure 4 medians by `r sprintf("%.0f", min(fig4_cmp$pct_diff))`% to
`r sprintf("%.0f", max(fig4_cmp$pct_diff))`% (median
`r sprintf("%.0f", median(fig4_cmp$pct_diff))`%). Each arm median from 100
virtual patients has a sampling error of about 6%, the 25 mg arms sit at
20-40 ng/mL where reading the figure is least precise, and the paper built
its cohorts by resampling individual predictions of its own patients rather
than drawing fresh random effects, so single-arm differences of this size
are expected. The immediate-release to sustained-release ratio of the
medians at 50 mg (`r sprintf("%.2f", ratio_ir_sr)`, sex-normalised) matches
the figure (98.9/75.0 = 1.32 in males, 147.3/110.8 = 1.33 in females).

```{r fig4-gate}
# Centre and ratio only (CLAUDE.md: never gate the extremes of a cohort).
# With 100 per arm the SE of a log-median is about 1.25 * 0.5 / 10 = 6%, so
# the IR/SR ratio of the 200-patient pooled medians has an SE near 0.06; the
# 1.05-1.60 window is more than 3 SE either side of the 1.26 typical-value
# ratio at day 30. A swapped formulation coding gives about 0.8, which fails.
# The median of the eight per-arm differences (about 6% on the authoring run,
# each arm with a ~6% sampling SE) is gated at 15%: a mis-transcribed CL/F,
# dose exponent or F moves every arm by 20% or more.
stopifnot(
  ratio_ir_sr > 1.05,
  ratio_ir_sr < 1.60,
  abs(median(fig4_cmp$pct_diff)) < 15
)
```

# PKNCA validation at steady state

Li 2022 reports no non-compartmental results, so the NCA check is internal:
PKNCA is run on the typical-value profiles over the 100th dosing interval
(2376-2400 h) and the average concentration is compared against the
closed-form steady-state value `F * dose / (CL/F * tau)`, which depends only on
the dose exponent, the sex effect on F and the clearance.

```{r pknca}
ev_ss <- make_events(typ_cohort, obs_times = c(0, seq(2376, 2400, by = 0.25)))
sim_ss <- rxode2::rxSolve(
  mod_typical,
  events = ev_ss,
  keep = c("DOSE", "SEXF", "FORM_PAROXETINE_IR")
) |>
  as.data.frame() |>
  dplyr::mutate(
    treatment = paste0(
      ifelse(FORM_PAROXETINE_IR == 1, "IR-T ", "SR-T "), DOSE, " mg, ",
      ifelse(SEXF == 1, "female", "male")
    )
  )

conc_df <- sim_ss |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(Cc = pmax(Cc, 0)) |>
  dplyr::select(id, time, Cc, treatment)

dose_df <- ev_ss |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, DOSE, SEXF, FORM_PAROXETINE_IR, amt, ii, addl) |>
  tidyr::uncount(addl + 1L, .id = "n") |>
  dplyr::mutate(
    time = (n - 1) * ii,
    treatment = paste0(
      ifelse(FORM_PAROXETINE_IR == 1, "IR-T ", "SR-T "), DOSE, " mg, ",
      ifelse(SEXF == 1, "female", "male")
    )
  ) |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)
intervals <- data.frame(
  start = 2376,
  end = 2400,
  cmax = TRUE,
  tmax = TRUE,
  cmin = TRUE,
  cav = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_res$result) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::left_join(
    sim_ss |> dplyr::distinct(treatment, DOSE, SEXF, FORM_PAROXETINE_IR),
    by = "treatment"
  ) |>
  dplyr::mutate(
    cav_closed = 1000 * (1 + 0.475 * SEXF) * DOSE / (21.2 * (DOSE / 40)^-1.03 * 24),
    pct_diff = 100 * (cav / cav_closed - 1)
  ) |>
  dplyr::arrange(FORM_PAROXETINE_IR, SEXF, DOSE)

nca_wide |>
  dplyr::select(treatment, cmax, tmax, cmin, cav, cav_closed, pct_diff) |>
  dplyr::rename(
    "Group" = treatment,
    "Cmax,ss (ng/mL)" = cmax,
    "Tmax (h after dose)" = tmax,
    "Cmin,ss (ng/mL)" = cmin,
    "Cav,ss PKNCA (ng/mL)" = cav,
    "Cav,ss closed form (ng/mL)" = cav_closed,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = 2, caption = "Steady-state NCA over the 100th dosing interval (typical values).")
```

```{r pknca-gate}
# The 0.25 h linear-up/log-down trapezoid on a profile this flat is accurate
# to well under 0.5%; residual accumulation at day 100 (largest for SR-T at
# 60 mg, half-life about 18 days, exp(-99 / 25.9) < 2.5%) is the only other
# contribution and it can only lower the PKNCA value.
stopifnot(
  nrow(nca_wide) == nrow(typ_cohort),
  all(nca_wide$pct_diff < 0.5),
  all(nca_wide$pct_diff > -4)
)
```

The steady-state peak-to-trough ratio is small for both tablets (half-lives
of days), which is why trough concentrations taken 10-22.5 h after the dose
are a good summary of exposure in this TDM population.

# Assumptions and deviations

- **Formulation coding.** The Methods code the immediate-release tablet as 1
  and the printed equation applies `(1 - 0.666 * formulation)` to V/F, giving
  the immediate-release tablet the smaller V/F (2956 L). The Abstract and
  Discussion describe the effect the other way round ("the SR-T resulted in a
  decrease in V/F of about 66.6%"). The Methods coding was kept because it is
  the one under which the paper's own typical-value simulations (Figure 3A
  and the peak-trough swings in Figure 5) are reproduced; under the
  Discussion's orientation the 720 h immediate-release troughs would differ
  from Figure 3A by +8% at 20 mg to -27% at 60 mg, and the Figure 5 swings
  would be exchanged.
- **Simulation horizon of the published figures.** The Methods state 100
  days of once-daily dosing, but Figures 3A and 5 are reproduced by the
  trough 24 h after the 30th dose. The library model is unaffected; this
  vignette compares at 720 h and also reports 2400 h.
- **Figure 4 cohorts.** The paper built Figure 4 by resampling individual
  predictions from its own patients, which cannot be reproduced without the
  data; this vignette draws fresh random effects from the reported IIV
  instead. Only the centre of the per-arm differences and the
  formulation ratio are gated, not single arms.
- **IIV scale.** Table 2 gives IIV as CV% without stating the conversion.
  The variances use `omega^2 = log(1 + CV^2)`. The alternative reading
  `omega^2 = (CV/100)^2` would give 0.269 (CL/F) and 0.697 (V/F); the paper
  reports no confidence intervals on the IIV rows that could separate the
  two readings.
- **Residual error scale.** Table 2 prints the proportional residual error
  as 'PRO (CV%) 0.0929' while the IIV rows in the same table are printed as
  percentages. The value was read as the raw NONMEM `$SIGMA` variance,
  giving a proportional SD of 0.305 (30.5%). If it were already an SD the
  residual error would be 9.3%; the residual error does not enter any of the
  typical-value or individual-prediction checks above.
- **Daily dose covariate.** `DOSE` is the total daily dose in mg. The paper
  centres it at 40 mg/day; the reported dose range is 20-75 mg/day, and the
  power model should not be extrapolated far outside it (at 10 mg/day the
  predicted CL/F would be 89 L/h).
- **Typographical error in the Results.** The Results state that males
  require "at least 5 mg QD" to reach a 20 ng/mL trough; with the model the
  typical male needs between 20 and 25 mg/day, consistent with the
  Discussion's statement that 20 mg/day in females corresponds to 25 mg/day
  in males.
- No erratum or correction notice for Li 2022 was indexed in Europe PMC
  as of 2026-10-09.
