---
title: "Non-Hodgkin lymphoma progression-free survival (Li 2017)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Non-Hodgkin lymphoma progression-free survival (Li 2017)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", error = FALSE,
  fig.width = 7, fig.height = 4.5
)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
mod <- rxode2::rxode2(readModelDb("Li_2017_nhl_pfs_mbma"))
```

* Citation: `r mod$reference`
* Article: <https://doi.org/10.1097/MD.0000000000007988>
* Supplemental Content (Model Description, Figure S1, Table S1):
  <http://links.lww.com/MD/B853>, also retrievable as
  `medi-96-e7988-s001.docx` from
  <https://www.ebi.ac.uk/europepmc/webservices/rest/PMC5585532/supplementaryFiles>

* Description: `r mod$description`

**The main text of Li 2017 contains no parameter table and no model
equation.** Both live in the Supplemental Content, which the Methods and
Results sections point to. Everything transcribed below comes from that
document; no value is taken from the abstract or from a figure.

# Population

The model was fitted to study-arm-level summary progression-free survival
(PFS) Kaplan-Meier curves digitised from published non-Hodgkin lymphoma
(NHL) trials. Of 513 NHL trials screened from PubMed, FDA reviews,
clinicaltrials.gov and the ASCO / ASH proceedings for 1993-2015, 179 met
the inclusion criteria, 17 were removed as duplicates, and 50 of the
remaining 162 were dropped for not reporting a PFS Kaplan-Meier curve.
The analysis dataset is therefore **112 studies, 155 cohorts (arms or
strata), 11,824 patients and 3,098 observations**, of which 22 studies
were randomised (Li 2017 Figure 1 and Results).

Cohort-level characteristics (Li 2017 Table 1): median age 63 years
across cohorts (range of cohort medians 47-83; not reported in 8 cohorts,
5.1%); median 59% male patients (range 29-91%; not reported in 14
cohorts, 9%). Histology strata: DLBCL 49 cohorts (31.6%), follicular
lymphoma 44 (28.4%), mantle cell lymphoma 33 (21.3%), mixed 25 (16.1%),
other 4 (2.5%). Treatment coverage, which overlaps because a cohort can
appear in several levels: rituximab 98 cohorts (63.2%), other
non-chemotherapy drugs 60 (38.7%), CHOP/CHOP-like 41 (26.4%), other
chemotherapy 36 (23.2%), bortezomib 22 (14.2%), bendamustine 11 (7%).

**No dose information was modelled.** Li 2017 Discussion states that
treatment duration, drug potency and dose "were not considered in the
model development, as there were not sufficient details available for
many of the trials", so every treatment covariate is a presence/absence
indicator. There is likewise no PK layer and no exposure metric.

The same information is available programmatically via the model's
`population` metadata (`readModelDb("Li_2017_nhl_pfs_mbma")()$population`).

# Source trace

Every `ini()` entry in
`inst/modeldb/therapeuticArea/oncology/Li_2017_nhl_pfs_mbma.R` carries an
in-file comment naming its source row. The table collects them.

| Equation / parameter | Value | Source location |
|---|---|---|
| Weibull survivor `PFS(t) = exp(-(t/lambda)^k)` | n/a | Supplemental Content, Model Description, equation (1) -- **as printed the equation is defective; see Errata** |
| Log-linear ("exponential") covariate model on `lambda` | n/a | Supplemental Content, Model Description ("Exponential model centered by median covariate value"; reference population = rituximab only / MCL / 100% treatment-naive) |
| Residual = scalar x SE of `ln(-ln(PFS(t)))` | n/a | Supplemental Content, Model Description (citing Arends 2008) |
| `lk` (shape `k`) | 0.79 | Table S1, Final Model, row *k* (RSE 4.3%, 95% CI 0.72, 0.85) |
| `llambda` (scale `lambda`, months) | 110 | Table S1, Final Model, row lambda (RSE 26.5%, 95% CI 52.8, 167.2) |
| `e_norituximab_llambda` | -1.13 | Table S1, Final Model, row "No rituximab on lambda" (RSE 9.4%, 95% CI -1.34, -0.92) |
| `e_chop_llambda` | 0.819 | Table S1, Final Model, row "CHOP/CHOP-like on lambda" (RSE 26.7%, 95% CI 0.39, 1.25) |
| `e_nonchemoother_llambda` | 0.813 | Table S1, Final Model, row "Other drugs on lambda" (RSE 8%, 95% CI 0.69, 0.94) |
| `e_bendamustine_llambda` | 1.39 | Table S1, Final Model, row "Bendamustine on lambda" (RSE 15.7%, 95% CI 0.96, 1.82) |
| `e_fl_llambda` | 0.46 | Table S1, Final Model, row "Follicular lymphoma on lambda" (RSE 25%, 95% CI 0.23, 0.68) |
| `e_dlbcl_llambda` | -1.41 | Table S1, Final Model, row "DLBCL on lambda" (RSE 8.5%, 95% CI -1.64, -1.18) |
| `e_othernhl_llambda` | -0.17 | Table S1, Final Model, row "Other lymphomas on lambda" (RSE 113%, 95% CI -0.54, 0.21) |
| `e_experienced_llambda` | -1.85 | Table S1, Final Model, row "% experienced patients on lambda" (RSE 12.8%, 95% CI -2.32, -1.39) |
| `eta_study_lk` (variance) | 0.17 | Table S1, Final Model, row omega^2_k (RSE 13.4%, shrinkage 5%) |
| `eta_study_llambda` (variance) | 1.4 | Table S1, Final Model, row omega^2_lambda (RSE 14.5%, shrinkage 2.3%) |
| `expSd` (= sqrt of the tabulated variance) | 1.0954451 | Table S1, Final Model, row sigma^2 = 1.2 (RSE 14.0%, shrinkage 3.4%) |
| Time unit = month | n/a | Li 2017 Figure 2 x-axis, "Time(month)" |
| Cohort characteristics in `population` | n/a | Li 2017 Table 1 and Results |
| Internal-validation percentages in `population` | n/a | Li 2017 Methods and Results |

Every tabulated value is checked against the loaded model below, so a
transcription slip fails the render rather than shipping.

```{r check-ini}
theta <- mod$theta
published <- c(
  lk                      = log(0.79),
  llambda                 = log(110),
  e_norituximab_llambda   = -1.13,
  e_chop_llambda          = 0.819,
  e_nonchemoother_llambda = 0.813,
  e_bendamustine_llambda  = 1.39,
  e_fl_llambda            = 0.46,
  e_dlbcl_llambda         = -1.41,
  e_othernhl_llambda      = -0.17,
  e_experienced_llambda   = -1.85,
  expSd                   = sqrt(1.2)
)
stopifnot(
  setequal(names(theta), names(published)),
  isTRUE(all.equal(theta[names(published)], published, tolerance = 1e-7))
)

# Table S1 reports omega^2 and sigma^2 as VARIANCES, so the diagonal of
# omega is used as tabulated and expSd is the square root of sigma^2.
stopifnot(
  isTRUE(all.equal(
    diag(mod$omega)[c("eta_study_lk", "eta_study_llambda")],
    c(eta_study_lk = 0.17, eta_study_llambda = 1.4),
    tolerance = 1e-12
  )),
  abs(theta[["expSd"]]^2 - 1.2) < 1e-7
)
```

# Structural model

PFS time follows a two-parameter Weibull distribution. In the
scale / shape parameterisation the Supplemental Content defines
(`lambda` the scale, `k` the shape, "also known as the Weibull slope"):

$$\mathrm{PFS}(t) = \exp\!\left[-\left(\frac{t}{\lambda}\right)^{k}\right],
\qquad
\ln\!\left(-\ln \mathrm{PFS}(t)\right) = k\ln t - k\ln\lambda .$$

The second form is the Weibull probability-plot linearisation, in which
`k` is literally the slope against `ln t` -- which is what the phrase
"Weibull slope" names, and which is the scale the residual error was
entered on. The model exposes the cumulative hazard
`cumhaz` $= (t/\lambda)^k$, the progression-free probability
`sur` $= \exp(-\mathtt{cumhaz})$, and the median PFS time

$$t_{\text{med}} = \lambda \,(\ln 2)^{1/k},$$

which is **directly proportional to `lambda`**. That proportionality is
why every covariate coefficient in Table S1 exponentiates to a
*median-PFS-time ratio*, and it is what makes the Li 2017 Figure 3 panels
an exact arithmetic check on the transcription rather than a visual one.

Three covariate blocks act log-linearly on `lambda`; **none** was
retained on `k` (Table S1 prints `---` for every "on *k*" row of the
final model). Two of the coefficients are tabulated with the *opposite*
polarity to the canonical covariate columns, which is the easiest thing
to get wrong in this model:

* Table S1 gives "**No** rituximab on lambda", so the coefficient is
  applied to `1 - CONMED_RITUX`.
* Table S1 gives "% **experienced** patients on lambda", so the
  coefficient is applied to `(100 - LINE_1L_PCT) / 100`.

Both follow from the Supplemental Content's reference population --
"rituximab as the only treatment, MCL as the tumor subtype, and 100%
treatment naive patients" -- and both are confirmed by Figure 3, where
MCL and "All naive patients" sit exactly on the ratio = 1.0 line.

Because the histology and line-of-therapy columns are canonical
*percentages* (0-100) while the fitted coefficients are per unit
*fraction*, the model divides each by 100.

```{r structure}
mod$covariates
```

# Validation

There is no NCA section in this vignette. The model has no concentration,
no dose and no PK layer, so `Cmax` / `AUC` / `t1/2` are undefined and
PKNCA is not the right instrument; the analogous checks for a parametric
survival model are the closed-form, monotonicity, median-recovery and
published-contrast gates below.

## Closed-form gate

The model is algebraic, so it can be checked against an independent R
implementation of the Weibull survivor evaluated at the **same drawn
between-study random effects**. Both sides use identical parameters, so
the only difference is numerical and a tight bound is the correct
assertion.

```{r closed-form}
rxode2::rxSetSeed(20260916)

grid_t <- c(0, seq(1, 120, by = 1))
n_cohort <- 200L

cohorts <- tibble(
  id = seq_len(n_cohort),
  CONMED_RITUX          = rep(c(1, 1, 1, 0), length.out = n_cohort),
  CONMED_CHOP           = rep(c(1, 0, 0, 0), length.out = n_cohort),
  CONMED_BENDAMUSTINE   = rep(c(0, 1, 0, 0), length.out = n_cohort),
  CONMED_NONCHEMO_OTHER = rep(c(0, 0, 1, 1), length.out = n_cohort),
  TUMTP_FL_PCT          = rep(c(100, 0, 0, 40), length.out = n_cohort),
  TUMTP_DLBCL_PCT       = rep(c(0, 100, 0, 30), length.out = n_cohort),
  TUMTP_OTHER_NHL_PCT   = rep(c(0, 0, 25, 10), length.out = n_cohort),
  LINE_1L_PCT           = rep(c(100, 0, 50, 25), length.out = n_cohort)
)

ev <- cohorts |> tidyr::expand_grid(time = grid_t)

sim <- rxode2::rxSolve(mod, ev)

# The drawn between-study etas live in $params, keyed by id; join them back so
# the closed form is evaluated at exactly the values the solver used.
etas <- as.data.frame(sim$params)[, c("id", "eta_study_lk", "eta_study_llambda")]
etas$id <- as.integer(as.character(etas$id))

closed <- as.data.frame(sim) |>
  mutate(id = as.integer(id)) |>
  left_join(etas, by = "id") |>
  mutate(
    k_ref = exp(published[["lk"]] + eta_study_lk),
    lam_ref = exp(
      published[["llambda"]] +
        published[["e_norituximab_llambda"]] * (1 - CONMED_RITUX) +
        published[["e_chop_llambda"]] * CONMED_CHOP +
        published[["e_bendamustine_llambda"]] * CONMED_BENDAMUSTINE +
        published[["e_nonchemoother_llambda"]] * CONMED_NONCHEMO_OTHER +
        published[["e_fl_llambda"]] * (TUMTP_FL_PCT / 100) +
        published[["e_dlbcl_llambda"]] * (TUMTP_DLBCL_PCT / 100) +
        published[["e_othernhl_llambda"]] * (TUMTP_OTHER_NHL_PCT / 100) +
        published[["e_experienced_llambda"]] * ((100 - LINE_1L_PCT) / 100) +
        eta_study_llambda
    )
  ) |>
  mutate(
    cumhaz_ref = (time / lam_ref)^k_ref,
    sur_ref = exp(-cumhaz_ref),
    tmed_ref = lam_ref * log(2)^(1 / k_ref)
  )

# Guard against a vacuous pass: a max() over zero rows silently returns -Inf,
# and a column typo would otherwise make every assertion below trivially true.
stopifnot(
  nrow(closed) == n_cohort * length(grid_t),
  all(c("k", "lambda", "cumhaz", "sur", "tmed",
        "eta_study_lk", "eta_study_llambda") %in% names(closed)),
  !anyNA(closed[, c("k", "lambda", "cumhaz", "sur", "tmed")])
)

stopifnot(
  max(abs(closed$k - closed$k_ref)) < 1e-12,
  max(abs(closed$lambda - closed$lam_ref)) < 1e-9,
  max(abs(closed$cumhaz - closed$cumhaz_ref)) < 1e-10,
  max(abs(closed$sur - closed$sur_ref)) < 1e-12,
  max(abs(closed$tmed - closed$tmed_ref)) < 1e-9
)
max(abs(closed$sur - closed$sur_ref))
```

## Survivor-function gates

A survivor function must start at exactly 1, decrease monotonically, stay
inside `[0, 1]`, and satisfy `sur = exp(-cumhaz)` exactly. These are the
checks that would have caught the mirror-image reading of the published
equation discussed under Errata: that reading gives an *increasing*
`sur`. The `sur(0) == 1` assertion is exact, and it is also what rules
out writing `cumhaz` with an epsilon offset on `t` -- see the note in
`model()` and the Assumptions section.

```{r survivor-gates}
stopifnot(
  # Starts at exactly 1: cumhaz(0) = 0^k = 0 for every k > 0.
  all(closed$cumhaz[closed$time == 0] == 0),
  all(closed$sur[closed$time == 0] == 1),
  # Bounded.
  all(closed$sur > 0), all(closed$sur <= 1),
  # sur = exp(-cumhaz) identically.
  max(abs(closed$sur - exp(-closed$cumhaz))) < 1e-15
)

# Strictly decreasing in time within every cohort.
mono <- closed |>
  arrange(id, time) |>
  group_by(id) |>
  summarise(max_increase = max(diff(sur)), .groups = "drop")
stopifnot(nrow(mono) == n_cohort, all(mono$max_increase < 0))
max(mono$max_increase)
```

The Weibull linearisation must hold exactly: regressing
`log(-log(sur))` on `log(time)` within a cohort must return slope `k` and
intercept `-k log(lambda)`.

```{r linearisation-gate}
lin <- closed |>
  filter(time > 0) |>
  group_by(id) |>
  summarise(
    slope = coef(lm(log(cumhaz) ~ log(time)))[[2]],
    intercept = coef(lm(log(cumhaz) ~ log(time)))[[1]],
    k = first(k),
    lambda = first(lambda),
    .groups = "drop"
  )
stopifnot(
  nrow(lin) == n_cohort,
  max(abs(lin$slope - lin$k)) < 1e-6,
  max(abs(lin$intercept + lin$k * log(lin$lambda))) < 1e-5
)
max(abs(lin$slope - lin$k))
```

The median PFS time must be the time at which `sur` crosses 0.5.

```{r median-gate}
tv <- rxode2::zeroRe(mod)

ref_cov <- list(
  CONMED_RITUX = 1, CONMED_CHOP = 0, CONMED_BENDAMUSTINE = 0,
  CONMED_NONCHEMO_OTHER = 0, TUMTP_FL_PCT = 0, TUMTP_DLBCL_PCT = 0,
  TUMTP_OTHER_NHL_PCT = 0, LINE_1L_PCT = 100
)

solve_tv <- function(overrides = list(), times = 0) {
  cv <- utils::modifyList(ref_cov, overrides)
  ev <- tibble(id = 1L, time = times)
  for (nm in names(cv)) ev[[nm]] <- cv[[nm]]
  as.data.frame(rxode2::rxSolve(tv, ev))
}

ref0 <- solve_tv()
tmed_ref <- ref0$tmed[[1]]
at_median <- solve_tv(times = tmed_ref)

stopifnot(
  # tmed = lambda * log(2)^(1/k).
  abs(tmed_ref - 110 * log(2)^(1 / 0.79)) < 1e-8,
  # and sur(tmed) = 0.5.
  abs(at_median$sur[[1]] - 0.5) < 1e-6
)
c(lambda = ref0$lambda[[1]], k = ref0$k[[1]], tmed_months = tmed_ref)
```

The reference cohort -- rituximab monotherapy, all mantle cell lymphoma,
all treatment-naive -- therefore has a median PFS time of
`r sprintf("%.1f", tmed_ref)` months.

## Covariate gate: the Li 2017 Figure 3 median-PFS-time ratios

This is the strongest available check on the transcription. Li 2017
Figure 3 plots each covariate's effect as a "median PFS time ratio"
against the reference population. Because `tmed` is proportional to
`lambda`, each ratio must equal `exp(coefficient)` exactly, and the
figure's own printed values pin the sign and the covariate scale
independently of Table S1.

```{r figure3-gate}
scenarios <- tribble(
  ~panel, ~label,                    ~overrides,                              ~expected,
  "3A",   "No rituximab",            list(CONMED_RITUX = 0),                   exp(-1.13),
  "3A",   "CHOP/CHOP-like added",     list(CONMED_CHOP = 1),                    exp(0.819),
  "3A",   "Bendamustine added",       list(CONMED_BENDAMUSTINE = 1),            exp(1.39),
  "3A",   "Other drugs added",        list(CONMED_NONCHEMO_OTHER = 1),          exp(0.813),
  "3B",   "All mantle cell (ref)",    list(),                                   1,
  "3B",   "All follicular lymphoma",  list(TUMTP_FL_PCT = 100),                 exp(0.46),
  "3B",   "All DLBCL",                list(TUMTP_DLBCL_PCT = 100),              exp(-1.41),
  "3B",   "All other NHL",            list(TUMTP_OTHER_NHL_PCT = 100),          exp(-0.17),
  "3C",   "All treatment-naive (ref)", list(LINE_1L_PCT = 100),                 1,
  "3C",   "All treatment-experienced", list(LINE_1L_PCT = 0),                   exp(-1.85)
)

ratios <- scenarios |>
  mutate(
    simulated = vapply(
      overrides,
      function(ov) solve_tv(ov)$tmed[[1]] / tmed_ref,
      numeric(1)
    ),
    abs_diff = abs(simulated - expected)
  )

stopifnot(
  nrow(ratios) == 10L,
  !anyNA(ratios$simulated),
  max(ratios$abs_diff) < 1e-12
)

ratios |>
  select(panel, label, simulated, expected) |>
  rename(
    "Figure 3 panel" = panel,
    "Cohort change from reference" = label,
    "Simulated median PFS time ratio" = simulated,
    "exp(published coefficient)" = expected
  ) |>
  knitr::kable(digits = 4)
```

The same numbers reproduce the quantitative claims Li 2017 makes in prose,
which are asserted here rather than merely narrated:

```{r prose-claims-gate}
get_ratio <- function(lbl) ratios$simulated[match(lbl, ratios$label)]

# Abstract: "Rituximab, bendamustine, CHOP/CHOP-like, and other
# nonchemotherapy drugs, aside from bortezomib, prolonged median PFS time
# 2 to 4-fold." The rituximab entry is the reciprocal of the tabulated
# no-rituximab effect.
fold <- c(
  rituximab    = 1 / get_ratio("No rituximab"),
  bendamustine = get_ratio("Bendamustine added"),
  chop         = get_ratio("CHOP/CHOP-like added"),
  other_drugs  = get_ratio("Other drugs added")
)
stopifnot(all(fold >= 2), all(fold <= 4.05))

# Abstract: "Follicular lymphoma patients had 60% longer median PFS time
# than mantle cell lymphoma (MCL) patients" -- 1.58, which rounds to the
# stated 60% at the paper's one-significant-figure precision.
stopifnot(abs(get_ratio("All follicular lymphoma") - 1.60) < 0.02)

# Abstract: "diffuse large B-cell lymphoma patients had a median PFS time
# that was 25% of MCL patients."
stopifnot(abs(get_ratio("All DLBCL") - 0.25) < 0.01)

# Results: other NHL subtypes were "similar to the MCL patient
# population" -- and the 95% CI of that coefficient brackets zero.
stopifnot(abs(get_ratio("All other NHL") - 1) < 0.16)

round(fold, 3)
```

## Replication of Li 2017 Figure 3

```{r figure3-plot, fig.height = 5}
ratios |>
  mutate(
    panel = recode(panel,
      "3A" = "A: treatment",
      "3B" = "B: NHL subtype",
      "3C" = "C: prior therapy"
    ),
    label = factor(label, levels = rev(label))
  ) |>
  ggplot(aes(x = simulated, y = label)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "blue") +
  geom_point(colour = "red", size = 2) +
  facet_wrap(~panel, ncol = 1, scales = "free_y") +
  scale_x_log10() +
  labs(
    x = "Median PFS time ratio (vs reference cohort)", y = NULL,
    caption = "Replicates Figure 3 of Li 2017 (point estimates only; the published confidence intervals are not reproducible from Table S1's coefficient CIs alone)"
  ) +
  theme_bw()
```

## Replication of Li 2017 Figure 2 (visual predictive check)

Figure 2 of Li 2017 is a visual predictive check of cohort-level %PFS
against time in months, with the observed 10th, 50th and 90th percentiles
across cohorts. The digitised observations are not publicly available, so
the panel below is the model's own predictive distribution over a virtual
database of 155 cohorts whose treatment, histology and prior-therapy mix
matches Li 2017 Table 1. It reproduces the qualitative shape of the
published figure: a steep early drop driven by the sub-1 Weibull shape,
followed by a long flat tail, with an extremely wide between-cohort
spread.

```{r vpc, fig.height = 4.5}
rxode2::rxSetSeed(20260916)

n_vpc <- 155L
set.seed(20260916)
vpc_cohorts <- tibble(
  id = seq_len(n_vpc),
  # Table 1 treatment coverage, as independent Bernoulli draws at the
  # published per-level cohort prevalences.
  CONMED_RITUX          = rbinom(n_vpc, 1, 0.632),
  CONMED_CHOP           = rbinom(n_vpc, 1, 0.264),
  CONMED_BENDAMUSTINE   = rbinom(n_vpc, 1, 0.070),
  CONMED_NONCHEMO_OTHER = rbinom(n_vpc, 1, 0.387)
) |>
  # Histology: assign each cohort a Table 1 stratum, then convert to the
  # percentage columns. "Mixed" cohorts get an even split.
  mutate(
    stratum = sample(
      c("FL", "MCL", "DLBCL", "OTHER", "MIXED"), n_vpc, replace = TRUE,
      prob = c(0.284, 0.213, 0.316, 0.025, 0.161)
    ),
    TUMTP_FL_PCT = case_when(
      stratum == "FL" ~ 100, stratum == "MIXED" ~ 100 / 3, TRUE ~ 0
    ),
    TUMTP_DLBCL_PCT = case_when(
      stratum == "DLBCL" ~ 100, stratum == "MIXED" ~ 100 / 3, TRUE ~ 0
    ),
    TUMTP_OTHER_NHL_PCT = case_when(stratum == "OTHER" ~ 100, TRUE ~ 0),
    # Table 1 "Number of treatments": 33.5% of cohorts on 1 regimen are
    # treated here as first-line, the rest as relapsed / refractory.
    LINE_1L_PCT = ifelse(rbinom(n_vpc, 1, 0.335) == 1, 100, 0)
  ) |>
  select(-stratum)

vpc_grid <- c(0, seq(1, 120, by = 1))
vpc_sim <- rxode2::rxSolve(
  mod, vpc_cohorts |> tidyr::expand_grid(time = vpc_grid)
) |>
  as.data.frame()

stopifnot(
  nrow(vpc_sim) == n_vpc * length(vpc_grid),
  !anyNA(vpc_sim$sur),
  all(vpc_sim$sur >= 0), all(vpc_sim$sur <= 1)
)

vpc_q <- vpc_sim |>
  group_by(time) |>
  summarise(
    p10 = quantile(sur, 0.10),
    p50 = quantile(sur, 0.50),
    p90 = quantile(sur, 0.90),
    .groups = "drop"
  )

ggplot(vpc_q, aes(x = time)) +
  geom_ribbon(aes(ymin = p10, ymax = p90), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = p50), colour = "red", linewidth = 1) +
  geom_line(aes(y = p10), colour = "blue") +
  geom_line(aes(y = p90), colour = "blue") +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "Time (month)", y = "Progression free survival (fraction)",
    caption = "Replicates the layout of Figure 2 of Li 2017: model-predicted median (red) and 10th / 90th percentiles (blue) of cohort-level PFS across a 155-cohort virtual database. Observed data are not publicly available."
  ) +
  theme_bw()
```

The predicted spread is deliberately wide: the between-study variance on
`log(lambda)` is 1.4, i.e. a standard deviation of
`r sprintf("%.2f", sqrt(1.4))` on the log scale, so the central 95% of
cohorts span a `r sprintf("%.0f", exp(4 * sqrt(1.4)))`-fold range in
`lambda` after the covariates have been accounted for. That is what the
Figure 2 percentile bands look like in the published paper too.

```{r vpc-spread-gate}
# The between-study spread is a structural property of omega, not of which
# cohorts happened to be drawn, so it is asserted on the model object
# rather than on the simulated sample.
stopifnot(
  abs(sqrt(diag(mod$omega)[["eta_study_llambda"]]) - sqrt(1.4)) < 1e-12,
  abs(sqrt(diag(mod$omega)[["eta_study_lk"]]) - sqrt(0.17)) < 1e-12
)

# The predicted median curve must be monotone and must stay inside the
# 10th-90th band; both are properties of the model, not of the draw.
stopifnot(
  all(diff(vpc_q$p50) <= 0),
  all(vpc_q$p50 >= vpc_q$p10), all(vpc_q$p50 <= vpc_q$p90)
)
```

## Hazard-ratio scope

Li 2017 also validated the model against published hazard ratios (100% of
cohorts within 0.5- to 2-fold of the observed HR, 75% within 25%). Under
a Weibull model two cohorts sharing the same shape `k` have a
*time-constant* hazard ratio equal to `(lambda_1 / lambda_2)^(-k)`, so a
between-arm HR is computable from this model directly. Because `k` itself
carries between-study variability, that identity only holds within a
study -- which is exactly the comparison a randomised trial's HR makes.

```{r hazard-ratio-gate}
hr_arms <- tibble(id = 1:2) |>
  mutate(
    CONMED_RITUX = 1, CONMED_CHOP = c(0, 1), CONMED_BENDAMUSTINE = 0,
    CONMED_NONCHEMO_OTHER = 0, TUMTP_FL_PCT = 0, TUMTP_DLBCL_PCT = 100,
    TUMTP_OTHER_NHL_PCT = 0, LINE_1L_PCT = 100
  )

hr_sim <- rxode2::rxSolve(
  rxode2::zeroRe(mod), hr_arms |> tidyr::expand_grid(time = c(6, 12, 24, 48))
) |>
  as.data.frame()

hr <- hr_sim |>
  select(id, time, cumhaz, k, lambda) |>
  tidyr::pivot_wider(
    names_from = id, values_from = c(cumhaz, k, lambda)
  ) |>
  mutate(
    # The cumulative-hazard ratio of two same-shape Weibulls is constant.
    hr_cumhaz = cumhaz_2 / cumhaz_1,
    hr_closed = (lambda_2 / lambda_1)^(-k_1)
  )

stopifnot(
  nrow(hr) == 4L,
  !anyNA(hr$hr_cumhaz),
  # Time-constant, and equal to the closed form.
  diff(range(hr$hr_cumhaz)) < 1e-9,
  max(abs(hr$hr_cumhaz - hr$hr_closed)) < 1e-9,
  # Adding CHOP to a rituximab backbone must REDUCE the hazard.
  all(hr$hr_cumhaz < 1)
)
hr |>
  select(time, hr_cumhaz) |>
  rename("Time (month)" = time, "Hazard ratio, R-CHOP vs rituximab alone" = hr_cumhaz) |>
  knitr::kable(digits = 4)
```

# Assumptions and deviations

* **Residual error is stored UNWEIGHTED.** The Supplemental Content
  states the residual was "the standard error of `ln(-ln(PFS(t)))`,
  scaled by an estimated constant". The per-observation residual SD is
  therefore `expSd * SE_i`, where `SE_i` is the standard error of the
  cohort's own transformed Kaplan-Meier estimate at that time point --
  a property of the digitised dataset (it depends on the number still at
  risk) that Li 2017 does not print a formula for. `expSd` in the model
  file is the unweighted scalar `sqrt(1.2)`; downstream simulation code
  must multiply it by `SE_i`. This follows the library's ratified
  convention for MBMA models whose observation weight touches only the
  residual (see the `N_ARM` register entry and the
  `Mercier_2014_tramadol_tapentadol_mbma` /
  `Chen_2025_methotrexate_*_mbma` precedent). Li 2017's between-study
  etas are *not* SE-weighted, so no weighting column is needed inside
  `model()`.
* **The residual is encoded as log-normal (`~ lnorm()`), not additive.**
  A residual additive on `ln(-ln(PFS))` is by construction additive on
  `ln(cumhaz)`, i.e. log-normal on `cumhaz`. `cumhaz ~ lnorm(expSd)` is
  therefore the source error model exactly, not an approximation of it.
* **Both between-study etas are encoded as exponential (log-scale).** The
  source says only that "inter-study variability was included on
  `lambda` and `k`" and the Table S1 footnote calls the numbers
  variances; it does not state the transform. The magnitudes force the
  log scale: an additive eta on `lambda` with SD `sqrt(1.4) = 1.18`
  *months* against a 110-month scale is a 1% between-study spread, which
  cannot coexist with the cohort spread in Figure 2, and an additive eta
  on `k` with SD `sqrt(0.17) = 0.41` against `k = 0.79` would put roughly
  3% of studies at an inadmissible `k < 0`. The log scale also matches
  the exponential covariate model these same parameters carry.
* **Reference-population centering, not median centering.** The
  Supplemental Content says the exponential covariate model was
  "centered by the median covariate value", but also states the
  reference population explicitly (rituximab only / MCL / 100%
  treatment-naive). The reference-population reading is implemented,
  because it is the only one under which Figure 3 places MCL and "All
  naive patients" exactly on the ratio = 1.0 line. The database medians
  are not reported for the retained covariates in any case.
* **`cumhaz` carries no epsilon offset on `t`, deliberately.** The usual
  library idiom for an algebraic model that takes a log of time is
  `((t + 1e-6) / lambda)^k`, so the log stays finite at `t = 0`. That
  idiom is wrong for a Weibull with a shape below 1, because a small
  offset raised to a *small power* is not small: at `k = 0.24` (well
  inside this model's between-study distribution, since
  `k = 0.79 * exp(eta)` with variance 0.17) and `lambda = 10` months,
  `(1e-6 / 10)^0.24 = 0.021`, a 2% drop in the survivor function at time
  zero. The exact form gives `0^k = 0` and `sur(0) = 1` identically.
  The price is that the *observation* scale `ln(cumhaz)` is `-Inf` at
  `t = 0`; that is faithful, because `ln(-ln(PFS))` is undefined wherever
  `PFS = 1` and the source's own transformed dataset therefore cannot
  contain a time-zero record either. Place observation records at
  `t > 0`. (`rxSolve()` floors the `lnorm` prediction at
  `.Machine$double.eps` rather than returning `NaN`, so a stray
  time-zero row does not poison a solve.)
* **Confidence intervals are not reproducible.** Table S1 gives a 95% CI
  per coefficient but no covariance matrix, so the Figure 3 error bars
  cannot be regenerated. The Figure 3 replication above plots point
  estimates only.
* **Covariate scale.** The canonical `_PCT` columns are percentages
  (0-100) while Li 2017's coefficients are per unit fraction; the model
  divides by 100. Li 2017 Table 1 prints its aggregate covariates as
  fractions in 0-1, so a user transcribing the source dataset directly
  must multiply by 100 before supplying these columns.
* **The prior-therapy covariate is linear in a column the paper treats
  as a threshold.** Li 2017's final model carries a single continuous
  "% experienced patients" coefficient, while its own subanalysis found
  treatment-naive and one-prior-line cohorts indistinguishable and
  two-or-more-prior-line cohorts at under a tenth the median PFS time.
  The linear form is the source model's; the subanalysis parameters are
  not tabulated anywhere and are not extracted.
* **The sensitivity analysis is not extracted.** Table S1's second block
  reports a complete-covariate sensitivity analysis on the 91 studies
  with no imputed covariates. Per the library's
  replicate-author-structure policy, robustness checks the authors did
  not report as final are excluded. For reference, it moved `k` to 0.69
  (95% CI 0.47, 0.91) and `lambda` to 118 (95% CI 35.7, 200.3) -- under
  15%, as the paper states -- but additionally retained treatment and
  tumor-subtype effects on `k` that the final model eliminated, and moved
  the bendamustine coefficient from 1.39 to 0.526 and the follicular-
  lymphoma coefficient from +0.46 to -0.142 (RSE 103%).
* **Bortezomib and other chemotherapy carry no coefficient.** Both were
  screened as their own treatment levels and eliminated in backward
  elimination; Table S1 prints `---` and Figure 3A places both exactly on
  the ratio = 1.0 line. A cohort receiving either is represented in this
  model by its other regimen components only. They are recorded in
  `covariatesDataExcluded` so the provenance of the screen is preserved.
* **Screened-and-dropped demographic covariates.** Median age, sex
  distribution, prior rituximab exposure, performance status >= 2 and
  disease stage >= III were all tested and none was significant at
  P < 0.01; none has a published point estimate. All are recorded in
  `covariatesDataExcluded`.
* **Simulation scope is study-arm level.** The random effects are
  between-STUDY, so the model simulates cohort-level PFS curves, median
  PFS times and between-arm hazard ratios. It does not simulate
  individual-patient event times, and a single solve is one *trial arm*,
  not one patient.
* **The virtual cohort in the Figure 2 replication is constructed, not
  published.** Li 2017 Table 1 gives marginal cohort prevalences for each
  treatment and histology level but not their joint distribution, so the
  155-cohort database above draws the treatment indicators independently
  at the published marginals and assigns histology strata by the
  published proportions. It is a plausibility panel, not a reproduction
  of the paper's own predictive distribution.

# Errata

**Supplemental Content, Model Description, equation (1) is defective as
published.** The supplement ships the equation as a 241 x 19 pixel raster
image (`word/media/image1.png` inside `medi-96-e7988-s001.docx`; the
document contains no OMML and no MathType object, so no
higher-resolution original exists). Magnified, that image reads

```
ln(-ln(PFS(t))) = k * ln(t) - k * ln(t)
```

whose right-hand side is identically zero. The `lambda` of the second
term was lost in typesetting. The intended equation is the standard
Weibull probability-plot linearisation

```
ln(-ln(PFS(t))) = k * ln(t) - k * ln(lambda)
```

equivalently `PFS(t) = exp(-(t/lambda)^k)`, which is what this model
implements. Three independent checks fix that reading and exclude the
mirror image `k*ln(lambda) - k*ln(t)`:

1. The mirror image gives a survivor function **increasing** in `t`
   (`PFS -> 1` as `t -> infinity`), which is inadmissible and contradicts
   the monotone-decreasing visual predictive check in Li 2017 Figure 2.
   The monotonicity gate above is the assertion that would fail.
2. The supplement's own text defines `lambda` as "the scale parameter"
   and `k` as "the shape parameter (also known as the Weibull slope)".
   "Weibull slope" names the coefficient of `ln t` in precisely this
   linearisation, so `k` must multiply `+ln(t)`.
3. Under this reading the median PFS time is `lambda * ln(2)^(1/k)`,
   proportional to `lambda`, which is what makes every Table S1
   coefficient exponentiate to the median-PFS-time ratio plotted in
   Figure 3. All ten Figure 3 entries reproduce to machine precision in
   the covariate gate above; under any other reading of the equation they
   would not.

No parameter value was tuned or back-solved to reach this conclusion.

**No erratum or corrigendum was located** for Li 2017 on the journal's
landing page or in PubMed; the only correction-shaped artefact is the
equation raster described above, which is a typesetting loss in the
supplement rather than a published correction.

**The supplement's backward-elimination narrative disagrees with its own
Table S1 on one word.** The Model Description says "Treatment
(bortezomib, other drugs) on lambda ... were removed in the backward
elimination steps", but Table S1's final-model column carries
"Other drugs on lambda" = 0.813 (RSE 8%, 95% CI 0.69 to 0.94), and both
the Abstract ("other nonchemotherapy drugs, aside from bortezomib,
prolonged median PFS time 2 to 4-fold") and Figure 3A place "Other drugs"
well above 1.0 and "Other chemo" exactly at 1.0. The parenthetical
evidently means "(bortezomib, other **chemotherapy**)". Table S1 and
Figure 3A govern, and the model retains the other-drugs effect and drops
bortezomib and other chemotherapy.
