---
title: "Amikacin (Li 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Amikacin (Li 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Li_2025_amikacin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Li_2025_amikacin"))$description`
* Article: <https://doi.org/10.1186/s12879-025-11747-z>
* Open-access full text: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC12797814/>

No supplementary material accompanies this article; every value below comes
from the main text, its five numbered equations, or Tables 1-5.

# Population

Li 2025 is a two-centre retrospective therapeutic-drug-monitoring study run in
the neonatal intensive care units of the Affiliated Suzhou Hospital of Nanjing
Medical University and the Children's Hospital of Soochow University (Suzhou,
Jiangsu, China), using clinical records from January 2021 to December 2022.
Twenty-three premature infants (gestational age < 37 weeks) treated with
amikacin for carbapenem-resistant-organism nosocomial pneumonia contributed 54
serum amikacin concentrations.

Baseline characteristics (Li 2025 Table 1): gestational age 28.90 +/- 2.53
weeks, postmenstrual age median 32.1 weeks (range 29.1-39.1), postnatal age
29.56 +/- 13.53 days, birth weight median 1.10 kg (range 0.70-4.30), weight at
amikacin administration median 1.36 kg (range 0.80-4.00), 13/23 (56.5%) male,
serum creatinine 31.72 +/- 11.06 umol/L and Schwartz eGFR 43.49 +/- 3.58
mL/min/1.73m^2. All 23 had CRO nosocomial pneumonia; 19 (82.6%) also had
bloodstream infection and 8 (34.8%) suppurative meningitis. Amikacin was given
as a 0.5-h intravenous infusion at a median of 14.32 mg/kg/day (range
10.34-19.70), usually 15 mg/kg/day split every 12 h, for a median of 12 days
(Li 2025 Table 2). Samples were drawn 1 h after the end of an infusion and 30
min before the next dose, all after the fifth dose, with additional random
inter-dose times; the LC-MS/MS assay LLOQ was 0.8 ug/mL.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Li_2025_amikacin")()$population`).

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Li_2025_amikacin.R` carries an in-file comment
naming its origin. They are collected here for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| One compartment, first-order elimination | n/a | Li 2025 Results, "Population PK analysis" |
| `d/dt(central) <- -kel * central` | n/a | Li 2025 Results, "Population PK analysis" |
| `cl` (Eq. 6) | `1.43 * (WT/70)^0.75 * fpma * rf * exp(etalcl)` | Li 2025 Eq. 6, p. 6 |
| `vc` (Eq. 7) | `30.97 * (WT/70)^1 * exp(etalvc)` | Li 2025 Eq. 7, p. 6 |
| `fpma` (Eq. 3) | `1 + SLPCL * (PMA - 40)` | Li 2025 Eq. 3, p. 4 |
| `rf` (Eq. 4) | `(CPR / Scr) / 6` | Li 2025 Eq. 4 + Eq. 2-5 legend, p. 4 (see Errata) |
| `cpr` (Eq. 5) | `516 * exp(Kage * ((PMA - 40)/52 - 40))` | Li 2025 Eq. 5, p. 4 |
| `lcl` | `log(1.43)` L/h per 70 kg | Li 2025 Table 4, tvCL (SE 0.14, CV 9.58%, 95% CI 1.15-1.70) |
| `lvc` | `log(30.97)` L per 70 kg | Li 2025 Table 4, tvV (SE 2.89, CV 9.32%, 95% CI 25.17-36.77) |
| `e_wt_cl` | `fixed(0.75)` | Li 2025 Eq. 2 legend, "PWR ... for CL is 0.75" |
| `e_wt_vc` | `fixed(1.0)` | Li 2025 Eq. 2 legend, "PWR ... V is 1" |
| `e_page_cl` (SLPCL) | `fixed(0.032)` per week | Li 2025 Table 4, SLPCL (no SE / CI / bootstrap) |
| `e_page_cpr` (Kage) | `fixed(0.00823)` per year | Li 2025 Table 4, Kage (no SE / CI / bootstrap) |
| `wt_ref` | 70 kg | Li 2025 Eq. 2 legend |
| `pma_ref` | 40 weeks | Li 2025 Eqs. 3 and 5 |
| `age_ref` | 40 years | Li 2025 Eq. 5 |
| `cpr_adult` | 516 umol/h per 70 kg | Li 2025 Methods, "the CPR was 516 umol.h-1" |
| `clcr_ref` | 6 L/h per 70 kg | Li 2025 Eq. 2-5 legend, "calibrated by CLcr of 6 L/h.per (70 kg)-1" |
| `etalcl` | `~ 0.16` (variance) | Li 2025 Table 4, omega^2 CL (SE 0.046, 95% CI 0.085-0.23, shrinkage 4.61%) |
| `etalvc` | `~ 0.15` (variance) | Li 2025 Table 4, omega^2 V (SE 0.047, 95% CI 0.073-0.23, shrinkage 10.35%) |
| `addSd` | 0.92 mg/L | Li 2025 Table 4, stdev0 (SE 0.10, CV 11.14%, 95% CI 0.72-1.13) |

Covariates screened by the stepwise covariate model but not retained (gender,
GA, PNA, birth weight, Apgar scores, haemoglobin, platelets, ALT, AST, total
and direct bilirubin, albumin, urea nitrogen, Schwartz eGFR, concomitant
ibuprofen) are recorded in the model file's `covariatesDataExcluded` list. Li
2025 reports no point estimate for any of them.

# Numeric equivalence with the published equations

Before any simulation, confirm the packaged model evaluates Li 2025 Eqs. 2-7
exactly. The hand computation below is written straight from the printed
equations; the model's own `cl` and `vc` must match it to machine precision.

```{r equivalence}
mod     <- rxode2::rxode(readModelDb("Li_2025_amikacin"))
mod_typ <- rxode2::zeroRe(mod)

# Cohort-median covariates (Li 2025 Table 1)
wt_med   <- 1.36    # kg, weight at amikacin administration
pma_med  <- 32.1    # weeks postmenstrual age
scr_mean <- 31.72   # umol/L serum creatinine

ev_one <- rxode2::et(amt = 14.32 / 2 * wt_med, dur = 0.5, cmt = "central") |>
  rxode2::et(seq(0, 12, by = 0.25), cmt = "central") |>
  as.data.frame() |>
  mutate(WT = wt_med, PAGE = pma_med / 4.35, CREAT = scr_mean)

sim_one <- rxode2::rxSolve(mod_typ, ev_one, returnType = "data.frame")

# Li 2025 Eqs. 2-7, transcribed directly
fsize_hand <- (wt_med / 70)^0.75                                   # Eq. 2
fpma_hand  <- 1 + 0.032 * (pma_med - 40)                           # Eq. 3
cpr_hand   <- 516 * exp(0.00823 * ((pma_med - 40) / 52 - 40))      # Eq. 5
rf_hand    <- (cpr_hand / scr_mean) / 6                            # Eq. 4 + legend
cl_hand    <- 1.43 * fsize_hand * fpma_hand * rf_hand              # Eq. 6
vc_hand    <- 30.97 * (wt_med / 70)                                # Eq. 7

equiv <- tibble::tibble(
  quantity  = c("Fsize (Eq. 2)", "F_PMA (Eq. 3)", "CPR (Eq. 5)",
                "Renal function (Eq. 4)", "CL (Eq. 6)", "V (Eq. 7)"),
  hand      = c(fsize_hand, fpma_hand, cpr_hand, rf_hand, cl_hand, vc_hand),
  model     = c(NA, NA, NA, NA, sim_one$cl[1], sim_one$vc[1])
)
knitr::kable(equiv, digits = 8,
             caption = "Hand evaluation of Li 2025 Eqs. 2-7 vs the packaged model.")

# Hard gate: the packaged model must reproduce the published equations exactly.
stopifnot(
  isTRUE(all.equal(sim_one$cl[1], cl_hand, tolerance = 1e-10)),
  isTRUE(all.equal(sim_one$vc[1], vc_hand, tolerance = 1e-10))
)
```

At the cohort-median covariates the model gives
CL = `r signif(cl_hand, 4)` L/h (`r signif(cl_hand / wt_med * 1000 / 60, 3)`
mL/min/kg), V = `r signif(vc_hand, 4)` L
(`r signif(vc_hand / wt_med, 3)` L/kg, consistent with amikacin's distribution
into extracellular fluid) and a terminal half-life of
`r signif(log(2) * vc_hand / cl_hand, 3)` h.

# Why the renal-function factor carries the 1/6 calibration

Printed Eq. 6 abbreviates the renal-function term to `CPR / Creatinine`, but
the Methods legend to Eqs. 2-5 states that renal function "was calibrated by
CLcr of 6 L/h per (70 kg)^-1". `CPR / Scr` has units
(umol/h per 70 kg) / (umol/L) = L/h per 70 kg, i.e. it *is* a creatinine
clearance, so the dimensionless factor multiplying CL must be `CLcr / 6`. The
calibration is self-consistent at adult reference values: 516 umol/h per 70 kg
divided by 86 umol/L (1.0 mg/dL, normal adult serum creatinine) is exactly
6.0 L/h per 70 kg, so the factor equals 1 in a normal adult.

Omitting the calibration inflates every clearance six-fold. The table below
contrasts the two readings against the observed concentrations Li 2025 reports
for its own cohort (Table 2: peak median 17.45 ug/mL, trough median 4.07
ug/mL), simulating the cohort-median subject on the most common regimen
(7.16 mg/kg every 12 h, 0.5-h infusion) and sampling at the study's own TDM
times.

```{r calibration-evidence}
# Steady-state profile for the cohort-median subject under both readings.
tau <- 12
ev_ss <- rxode2::et(amt = 14.32 / 2 * wt_med, dur = 0.5, ii = tau,
                    until = tau * 11, cmt = "central") |>
  rxode2::et(seq(tau * 11, tau * 12, by = 0.05), cmt = "central") |>
  as.data.frame() |>
  mutate(WT = wt_med, PAGE = pma_med / 4.35, CREAT = scr_mean)

sim_cal <- rxode2::rxSolve(mod_typ, ev_ss, returnType = "data.frame") |>
  filter(time >= tau * 11) |>
  mutate(tad = time - tau * 11)

# The uncalibrated reading multiplies CL by 6 and leaves V untouched.
kel_cal   <- cl_hand / vc_hand
kel_uncal <- 6 * cl_hand / vc_hand
prof <- function(kel, dose, tau, tad) {
  # Analytic steady-state 0.5-h infusion profile, used only for this contrast.
  r <- dose / 0.5
  cinf <- r / (kel * vc_hand) * (1 - exp(-kel * pmin(tad, 0.5))) /
    (1 - exp(-kel * tau))
  ifelse(tad <= 0.5, cinf,
         r / (kel * vc_hand) * (1 - exp(-kel * 0.5)) / (1 - exp(-kel * tau)) *
           exp(-kel * (tad - 0.5)))
}
dose_med   <- 14.32 / 2 * wt_med
tad_eoi    <- 0.5           # end of the 0.5-h infusion (the true peak)
tad_tdm    <- 1.5           # 1 h after the end of the infusion (the TDM sample)
tad_trough <- tau - 0.5     # 30 min before the next dose (the TDM trough)

calib <- tibble::tibble(
  Reading = c("With 1/6 calibration (packaged model)",
              "Without calibration (Eq. 6 as literally printed)"),
  `CL (L/h)` = c(cl_hand, 6 * cl_hand),
  `t1/2 (h)` = c(log(2) / kel_cal, log(2) / kel_uncal),
  `Peak at end of infusion (ug/mL)` =
    c(prof(kel_cal, dose_med, tau, tad_eoi), prof(kel_uncal, dose_med, tau, tad_eoi)),
  `Peak at TDM time, +1 h (ug/mL)` =
    c(prof(kel_cal, dose_med, tau, tad_tdm), prof(kel_uncal, dose_med, tau, tad_tdm)),
  `Trough (ug/mL)` =
    c(prof(kel_cal, dose_med, tau, tad_trough), prof(kel_uncal, dose_med, tau, tad_trough))
)
knitr::kable(calib, digits = 3,
             caption = paste("Cohort-median subject, 7.16 mg/kg q12h.",
                             "Li 2025 Table 2 observed medians:",
                             "peak 17.45 ug/mL, trough 4.07 ug/mL."))

# The discriminating quantities are the two TDM samples, not the peak at the
# end of the infusion: a very fast kel simply drives a 0.5-h infusion towards a
# plateau, so the end-of-infusion peak stays within about 30% either way. This
# is the evidence, not a tuning step.
rel <- function(x) abs(x - 17.45) / 17.45
stopifnot(
  # Calibrated: essentially exact at the end of the infusion, and within 20% at
  # the TDM peak sampling time 1 h later.
  rel(prof(kel_cal, dose_med, tau, tad_eoi)) < 0.05,
  rel(prof(kel_cal, dose_med, tau, tad_tdm)) < 0.20,
  # Uncalibrated: the TDM peak is more than 60% below the observed median and
  # the trough vanishes, against an observed trough median of 4.07 ug/mL.
  rel(prof(kel_uncal, dose_med, tau, tad_tdm))  > 0.60,
  prof(kel_uncal, dose_med, tau, tad_trough)    < 0.01,
  prof(kel_cal,   dose_med, tau, tad_trough)    > 1.5
)
```

Under the calibrated reading the peak at the end of the infusion is 17.5 ug/mL
against an observed median of 17.45 - essentially exact - and 14.6 ug/mL at the
paper's actual TDM sampling time 1 h later, 16% below the observed median. That
16% gap is the same "clearance runs fast" finding discussed in the Errata: it
is the extra hour of decay between the two conventions that costs the
agreement, not the volume of distribution.

The literal reading of Eq. 6 fails on both TDM samples. Its end-of-infusion
peak is 12.5 ug/mL, only about 30% below the observed median - a very fast
elimination rate drives a 0.5-h infusion towards a plateau, so that instant
alone is not decisive. But by the actual peak sampling time 1 h later the
concentration has fallen to 4.2 ug/mL, and the trough is zero to three decimal
places against an observed median of 4.07. A half-life of 0.64 h in a preterm
neonate is incompatible with the study's own observations and with its Table 5
simulations, so the packaged model uses the calibrated form the Methods legend
specifies.

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the simulation grid Li 2025 used for Table 5, restricted to the postmenstrual
age / weight combination reported in every creatinine band (PMA 31 weeks,
WT 1.2 kg). Serum creatinine is drawn uniformly inside each band, matching the
paper's "random distribution of creatinine level within a quartiles range".

The arms are derived from the published table itself, so every simulated arm
has a reference row and no arm is simulated without one.

```{r published-table5}
# Li 2025 Table 5, the PMA 31 wk / WT 1.2 kg rows of all three Scr bands.
# Percentages of simulated subjects by trough and peak band.
published <- tibble::tribble(
  ~scr_band, ~scr_lo, ~scr_hi, ~regimen,        ~mgkg, ~tau, ~trough_lt5, ~peak_lt20, ~peak_20_35, ~peak_gt35,
  "15-22",   15,      22,      "10 mg/kg q24h", 10,    24,   99.5,        38.6,       58.8,        2.6,
  "15-22",   15,      22,      "11 mg/kg q24h", 11,    24,   99.5,        28.0,       66.1,        5.9,
  "15-22",   15,      22,      "12 mg/kg q24h", 12,    24,   99.4,        19.6,       72.0,        8.4,
  "15-22",   15,      22,      "12 mg/kg q36h", 12,    36,   99.9,        21.8,       70.3,        8.0,
  "15-22",   15,      22,      "12 mg/kg q48h", 12,    48,  100.0,        22.1,       69.9,        8.0,
  "15-22",   15,      22,      "13 mg/kg q24h", 13,    24,   99.5,        15.0,       73.9,       11.1,
  "23-36",   23,      36,      "10 mg/kg q24h", 10,    24,   92.5,        26.6,       67.6,        5.8,
  "23-36",   23,      36,      "11 mg/kg q24h", 11,    24,   90.7,        19.3,       71.5,        9.2,
  "23-36",   23,      36,      "12 mg/kg q24h", 12,    24,   88.1,        13.1,       73.0,       13.9,
  "23-36",   23,      36,      "12 mg/kg q36h", 12,    36,   98.7,        16.9,       72.2,       10.9,
  "23-36",   23,      36,      "12 mg/kg q48h", 12,    48,   99.7,        17.8,       71.9,       10.3,
  "23-36",   23,      36,      "13 mg/kg q24h", 13,    24,   86.2,         8.8,       71.9,       19.4,
  "37-60",   37,      60,      "10 mg/kg q24h", 10,    24,   53.5,        14.7,       72.9,       12.5,
  "37-60",   37,      60,      "11 mg/kg q24h", 11,    24,   48.4,         9.7,       71.2,       19.1,
  "37-60",   37,      60,      "12 mg/kg q24h", 12,    24,   44.2,         6.3,       67.7,       26.0,
  "37-60",   37,      60,      "12 mg/kg q36h", 12,    36,   82.9,        11.1,       71.9,       17.0,
  "37-60",   37,      60,      "12 mg/kg q48h", 12,    48,   96.2,        14.1,       71.5,       14.3
)
```

Every arm is simulated with **common random numbers**: virtual subject *j* gets
the same pair of random effects and the same creatinine *quantile* within its
band in every arm. Target-attainment percentages are then compared pairwise
across regimens and creatinine bands, so the differences below reflect the dose
and the covariate, not Monte Carlo noise. Without this, a 120-subject arm
carries roughly +/- 4 percentage points of independent noise, which is larger
than several of the dose steps Li 2025 distinguishes.

```{r cohort}
crn_seed  <- 20251118L
n_per_arm <- 120L    # <= 200 per arm
pma_sim   <- 31      # weeks, Li 2025 Table 5
wt_sim    <- 1.2     # kg,    Li 2025 Table 5
n_dose    <- 8L      # enough intervals to reach steady state at every tau

arms <- published |>
  mutate(arm = sprintf("Scr %s | %s", scr_band, regimen),
         id_offset = (row_number() - 1L) * n_per_arm)

# One common vector of creatinine quantiles, reused by every arm.
set.seed(crn_seed)
creat_u <- runif(n_per_arm)

make_arm <- function(scr_lo, scr_hi, mgkg, tau, arm, id_offset) {
  t_last <- tau * (n_dose - 1L)
  subj <- tibble::tibble(
    id    = id_offset + seq_len(n_per_arm),
    WT    = wt_sim,
    PAGE  = pma_sim / 4.35,
    CREAT = scr_lo + creat_u * (scr_hi - scr_lo),
    arm   = arm,
    tlast = t_last
  )
  doses <- tidyr::crossing(subj, tibble::tibble(time = (seq_len(n_dose) - 1L) * tau)) |>
    mutate(amt = mgkg * wt_sim, evid = 1L, dur = 0.5, cmt = "central")
  # Observation grid over the final dosing interval only: dense through the
  # infusion and distribution phase, hourly afterwards, plus the exact TDM
  # sampling times (end of infusion, and 30 min before the next dose).
  # `cmt` is the ODE STATE name, never the algebraic observable `Cc`.
  grid <- sort(unique(c(seq(0, 4, by = 0.1), seq(4, tau, by = 1), 0.5, tau - 0.5, tau)))
  obs <- tidyr::crossing(subj, tibble::tibble(time = t_last + grid)) |>
    mutate(amt = NA_real_, evid = 0L, dur = NA_real_, cmt = "central")
  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}

events <- do.call(bind_rows, lapply(seq_len(nrow(arms)), function(i) {
  a <- arms[i, ]
  make_arm(a$scr_lo, a$scr_hi, a$mgkg, a$tau, a$arm, a$id_offset)
}))

# No two rows may share (id, time, evid) -- a duplicated dose would double the
# amount delivered, and a duplicated observation would double-weight that time
# in the NCA. Note the `unique()` must NOT be applied before `anyDuplicated()`:
# de-duplicating first makes the test vacuous (it can then never fail).
stopifnot(!anyDuplicated(events[, c("id", "time", "evid")]))
stopifnot(dplyr::n_distinct(events$id) == nrow(arms) * n_per_arm)
```

# Simulation

Each arm is solved separately with the random-number stream reset beforehand,
so subject *j* draws the same `etalcl` / `etalvc` pair in every arm.

```{r simulate}
solve_arm <- function(arm_name) {
  ev <- events |> filter(arm == arm_name)
  set.seed(crn_seed)
  rxode2::rxSetSeed(crn_seed)
  rxode2::rxSolve(mod, events = ev, keep = c("arm", "tlast")) |>
    as.data.frame()
}

sim <- do.call(bind_rows, lapply(arms$arm, solve_arm)) |>
  filter(!is.na(Cc)) |>
  mutate(tad = time - tlast)

stopifnot(all(sim$Cc >= 0))

# Confirm the common random numbers actually landed: subject j must have the
# same individual V in every arm (V depends only on WT and etalvc, both common).
vc_by_slot <- sim |>
  group_by(id) |>
  summarise(arm = first(arm), vc = first(vc), .groups = "drop") |>
  left_join(arms |> select(arm, id_offset), by = "arm") |>
  mutate(slot = id - id_offset)
stopifnot(
  vc_by_slot |>
    group_by(slot) |>
    summarise(spread = diff(range(vc)), .groups = "drop") |>
    pull(spread) |>
    max() < 1e-12
)
```

`Cc` is the individual prediction; the additive residual error is not added,
matching a model-based dosing simulation.

# Replicate Figure 7 - covariate effects on the concentration-time profile

Li 2025 Figure 7 shows typical-value predictions for 12 mg/kg q36h at
WT = 1.8 kg: panel A varies serum creatinine across the three bands at
PMA = 34 weeks, panel B varies PMA (31 vs 34 weeks) at Scr = 36 umol/L.

```{r figure-7}
fig7_profile <- function(pma_wk, wt, scr, mgkg, tau, nd = 8L) {
  t_last <- tau * (nd - 1L)
  ev <- rxode2::et(amt = mgkg * wt, dur = 0.5, ii = tau,
                   until = t_last, cmt = "central") |>
    rxode2::et(t_last + seq(0, tau, by = 0.1), cmt = "central") |>
    as.data.frame() |>
    mutate(WT = wt, PAGE = pma_wk / 4.35, CREAT = scr)
  rxode2::rxSolve(mod_typ, ev, returnType = "data.frame") |>
    filter(time >= t_last) |>
    mutate(tad = time - t_last)
}

scr_mids <- c(18.5, 29.5, 48.5)   # midpoints of the three Table 5 bands

fig7a <- bind_rows(lapply(scr_mids, function(s) {
  fig7_profile(34, 1.8, s, 12, 36) |>
    mutate(panel = "A: PMA 34 wk, varying Scr",
           grp = sprintf("Scr %.1f umol/L", s))
}))
fig7b <- bind_rows(lapply(c(31, 34), function(p) {
  fig7_profile(p, 1.8, 36, 12, 36) |>
    mutate(panel = "B: Scr 36 umol/L, varying PMA",
           grp = sprintf("PMA %d wk", p))
}))

bind_rows(fig7a, fig7b) |>
  ggplot(aes(tad, Cc, colour = grp)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~panel) +
  labs(x = "Time after dose (h)", y = "Amikacin (ug/mL)", colour = NULL,
       title = "Figure 7 - 12 mg/kg q36h, WT 1.8 kg, typical values",
       caption = "Replicates Figure 7 of Li 2025.")
```

The two covariate directions Li 2025 relies on for its dosing recommendation
are asserted rather than eyeballed: higher serum creatinine must raise the
trough (slower elimination) and higher postmenstrual age must lower it.

```{r figure-7-assertions}
trough_of <- function(df) df$Cc[which.max(df$tad)]

tr_scr <- vapply(scr_mids,
                 function(s) trough_of(fig7_profile(34, 1.8, s, 12, 36)), numeric(1))
tr_pma <- vapply(c(31, 34),
                 function(p) trough_of(fig7_profile(p, 1.8, 36, 12, 36)), numeric(1))

knitr::kable(
  tibble::tibble(
    Panel  = c(rep("A (Scr, PMA 34 wk)", 3), rep("B (PMA, Scr 36 umol/L)", 2)),
    Level  = c("Scr 18.5", "Scr 29.5", "Scr 48.5", "PMA 31 wk", "PMA 34 wk"),
    `36 h trough (ug/mL)` = c(tr_scr, tr_pma)
  ),
  digits = 4,
  caption = "Figure 7 trough concentrations: monotone in Scr and in PMA."
)

stopifnot(
  length(tr_scr) == 3L, all(diff(tr_scr) > 0),   # trough rises with creatinine
  length(tr_pma) == 2L, diff(tr_pma) < 0         # trough falls with PMA
)
```

# PKNCA validation

Steady-state NCA over the final dosing interval, grouped by simulation arm.

```{r pknca}
sim_nca <- sim |>
  select(id, arm, time = tad, Cc) |>
  arrange(arm, id, time) |>
  as.data.frame()

# Time zero of the interval is present by construction (the grid starts at 0),
# so no defensive row is needed; assert it rather than assume it.
stopifnot(all(
  sim_nca |> group_by(id) |> summarise(has0 = any(time == 0), .groups = "drop") |>
    pull(has0)
))

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id,
                             concu = "ug/mL", timeu = "h")

dose_df <- events |>
  filter(evid == 1) |>
  group_by(id) |>
  slice_max(time, n = 1) |>
  ungroup() |>
  transmute(id, arm, amt, time = 0) |>
  as.data.frame()

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id, doseu = "mg")

intervals <- arms |>
  transmute(arm, start = 0, end = tau,
            cmax = TRUE, tmax = TRUE, auclast = TRUE) |>
  as.data.frame()

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))

nca_tab <- as.data.frame(nca_res) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "auclast")) |>
  group_by(arm, PPTESTCD) |>
  summarise(med = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = med) |>
  rename("Arm" = arm, "Cmax (ug/mL)" = cmax, "Tmax (h)" = tmax,
         "AUC0-tau (ug*h/mL)" = auclast)

knitr::kable(nca_tab, digits = 2,
             caption = "Median steady-state NCA per simulation arm.")
```

## Structural identity: AUC0-tau = Dose / CL per subject

At steady state the area under one dosing interval equals dose divided by the
individual's clearance. This is an exact per-subject identity and is the
strongest available check that the simulation, the NCA window and the model all
agree.

```{r auc-identity}
cl_by_id <- sim |> group_by(id) |> summarise(cl = first(cl), .groups = "drop")

auc_check <- as.data.frame(nca_res) |>
  filter(PPTESTCD == "auclast") |>
  select(id, arm, auc = PPORRES) |>
  left_join(cl_by_id, by = "id") |>
  left_join(dose_df |> select(id, amt), by = "id") |>
  mutate(auc_expected = amt / cl, rel_err = auc / auc_expected - 1)

knitr::kable(
  auc_check |>
    summarise(n = n(),
              `median rel. error` = median(rel_err),
              `max |rel. error|`  = max(abs(rel_err))),
  digits = 5,
  caption = "AUC0-tau vs Dose/CL, over every simulated subject."
)

stopifnot(nrow(auc_check) == nrow(arms) * n_per_arm)
# The worst per-subject trapezoidal error on this observation grid is 1.14%
# (a fast-clearing subject on the hourly part of the grid), so 1.5% is what is
# asserted -- tight enough to catch a regression, not merely to pass.
stopifnot(max(abs(auc_check$rel_err)) < 0.015)
```

## Comparison against Li 2025 Table 5

Peaks are read at the end of the 0.5-h infusion and troughs 30 min before the
next dose, matching the study's own TDM sampling times.

```{r table5-compare}
tau_by_arm <- arms |> select(arm, tau)

attain <- sim |>
  left_join(tau_by_arm, by = "arm") |>
  group_by(id) |>
  summarise(
    arm    = first(arm),
    # End of the 0.5-h infusion, and 30 min before the next dose: the study's
    # own TDM sampling times. Both are exact points on the observation grid.
    peak   = Cc[which.min(abs(tad - 0.5))],
    trough = Cc[which.min(abs(tad - (first(tau) - 0.5)))],
    .groups = "drop"
  ) |>
  group_by(arm) |>
  summarise(
    sim_trough_lt5 = 100 * mean(trough < 5),
    sim_peak_lt20  = 100 * mean(peak < 20),
    sim_peak_20_35 = 100 * mean(peak >= 20 & peak <= 35),
    sim_peak_gt35  = 100 * mean(peak > 35),
    .groups = "drop"
  )

cmp5 <- arms |>
  left_join(attain, by = "arm")

stopifnot(!anyNA(cmp5$sim_trough_lt5))   # every published row got a simulated arm

cmp5 |>
  transmute(
    "Scr (umol/L)"          = scr_band,
    "Regimen"               = regimen,
    "Trough <5, paper (%)"  = trough_lt5,
    "Trough <5, sim (%)"    = sim_trough_lt5,
    "Peak 20-35, paper (%)" = peak_20_35,
    "Peak 20-35, sim (%)"   = sim_peak_20_35,
    "Peak >35, paper (%)"   = peak_gt35,
    "Peak >35, sim (%)"     = sim_peak_gt35
  ) |>
  knitr::kable(digits = 1,
               caption = paste("Li 2025 Table 5 (PMA 31 wk, WT 1.2 kg) vs the",
                               "packaged model. No parameter was adjusted."))
```

The simulation reproduces the structure Li 2025 relies on but not the
percentages. Two systematic gaps show up. Simulated trough *concentrations*
sit below Table 5's, so the fraction below 5 ug/mL comes out correspondingly
*higher* - most visibly in the worst renal band, where the model puts 86% of
subjects under 5 ug/mL on 10 mg/kg q24h against Table 5's 53.5%. And the
simulated peak distribution is wider, so the fraction landing inside the
20-35 ug/mL window is lower (about 40-63% against Table 5's near-uniform
58-74%) with fatter tails on both sides. Both gaps are recorded in the Errata
below; no parameter was adjusted to close them.

What is reproducible is the mechanism the paper's dosing recommendation rests
on. Three claims from the Results and Discussion are asserted directly.

```{r table5-claims}
tr <- function(band, reg) cmp5$sim_trough_lt5[cmp5$scr_band == band & cmp5$regimen == reg]
pk <- function(band, reg) cmp5$sim_peak_gt35[cmp5$scr_band == band & cmp5$regimen == reg]
one <- function(x) { stopifnot(length(x) == 1L); x }   # a lookup that misses must fail loudly

# Claim 1 (Discussion): "Scr significantly affected the elimination of
# amikacin" - at a fixed regimen, trough attainment falls as Scr rises.
c1 <- vapply(c("15-22", "23-36", "37-60"), function(b) one(tr(b, "12 mg/kg q24h")), numeric(1))

# Claim 2 (Discussion): the optimised regimens are "gradually reduced from
# 13 mg/kg q24h to 12 mg/kg q36h or 48 h, with Scr increased from 15 to 60
# umol/L" - within the worst renal band, lengthening the interval at a fixed
# mg/kg restores trough attainment.
c2 <- vapply(c("12 mg/kg q24h", "12 mg/kg q36h", "12 mg/kg q48h"),
             function(r) one(tr("37-60", r)), numeric(1))

# Claim 3 (Results): overshoot of the 35 ug/mL peak ceiling rises with the
# mg/kg dose at a fixed 24 h interval.
c3 <- vapply(c("10 mg/kg q24h", "11 mg/kg q24h", "12 mg/kg q24h", "13 mg/kg q24h"),
             function(r) one(pk("23-36", r)), numeric(1))

knitr::kable(
  tibble::tibble(
    Claim = c(rep("1: trough attainment falls as Scr rises (12 mg/kg q24h)", 3),
              rep("2: lengthening tau restores trough attainment (Scr 37-60)", 3),
              rep("3: peak overshoot rises with mg/kg (Scr 23-36, q24h)", 4)),
    Level = c("Scr 15-22", "Scr 23-36", "Scr 37-60",
              "q24h", "q36h", "q48h",
              "10 mg/kg", "11 mg/kg", "12 mg/kg", "13 mg/kg"),
    `Simulated (%)` = c(c1, c2, c3)
  ),
  digits = 1,
  caption = "Assertions on the paper's own mechanistic claims."
)

stopifnot(
  all(diff(c1) < 0),   # trough attainment strictly falls as Scr rises
  all(diff(c2) > 0),   # strictly rises as the dosing interval lengthens
  all(diff(c3) > 0)    # peak overshoot strictly rises with the mg/kg dose
)
```

For reference, the regimen that maximises joint trough-and-peak attainment in
this simulation is shown against the paper's recommendation. Because the
simulated troughs are systematically lower than Table 5's, the trough term
discriminates less than it does in the paper, so the two need not agree; this
table is reported, not asserted.

```{r recommendation-compare}
# Li 2025 Conclusions: "13 mg/kg q24h, 12 mg/kg q36h and q48h for serum
# creatinine between 15-22 umol/L; 23-36 and 37-60 umol/L, respectively."
li_recommendation <- tibble::tribble(
  ~scr_band, ~recommended,
  "15-22",   "13 mg/kg q24h",
  "23-36",   "12 mg/kg q36h",
  "37-60",   "12 mg/kg q48h"
)

cmp5 |>
  group_by(scr_band) |>
  slice_max(sim_trough_lt5 / 100 * sim_peak_20_35 / 100, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(scr_band, sim_best = regimen) |>
  # Join by key, never by row position.
  left_join(li_recommendation, by = "scr_band") |>
  transmute(
    "Scr (umol/L)"           = scr_band,
    "Simulated best regimen" = sim_best,
    "Li 2025 recommendation" = recommended
  ) |>
  knitr::kable(caption = "Best regimen per creatinine band.")
```

# Assumptions and deviations

## Errata and resolved discrepancies in the source

* **Eq. 6 omits the renal-function calibration that the Methods legend
  states.** As printed, Eq. 6 multiplies clearance by `CPR / Creatinine`, an
  absolute creatinine clearance in L/h per 70 kg, which is about 6 at normal
  adult renal function and about 12 in this cohort. Taken literally, the
  typical infant's half-life would be under an hour and the trough
  indistinguishable from zero, contradicting the study's own observed troughs
  (median 4.07 ug/mL) and its Table 5 simulations. The Methods legend to
  Eqs. 2-5 supplies the missing piece - renal function "was calibrated by CLcr
  of 6 L/h per (70 kg)^-1" - and 516 / 86 = 6.0 exactly, so the packaged model
  uses `(CPR / Scr) / 6`. The "Why the renal-function factor carries the 1/6
  calibration" section above shows the numeric evidence.

* **Eq. 5 is typeset with unbalanced brackets.** The PDF renders it as
  `CPR = [516 * exp(Kage * [(PMA - 40)/52 - 40]`. The fraction bar places 52
  under `(PMA - 40)` with the `- 40` outside it, so the argument is age in
  years relative to term birth, `(PMA - 40)/52`, minus the 40-year adult
  reference age at which CPR equals 516 umol/h per 70 kg. That is the reading
  encoded, and it is the only one under which the model reproduces the
  observed peak concentrations.

* **`SLPCL` and `Kage` are encoded as `fixed()`.** Li 2025 Table 4 reports
  both as bare point estimates with "-" in every uncertainty column (SE, CV,
  95% CI) and omits them from the bootstrap columns, whereas tvCL, tvV, both
  omegas and stdev0 all carry a full SE / CV / 95% CI and a bootstrap median.
  That reporting split is the evidence they were held constant rather than
  estimated. `Kage` is in addition a constant of the creatinine-production
  model that Li 2025 attributes to its reference 30 (Allegaert 2008). Note
  that Table 3 reports a P-value for Model 1, which adds only the allometric
  exponents - values the paper explicitly calls "empirical coefficients" -
  so a Table 3 P-value is not evidence that a parameter was estimated. The
  distinction is provenance only; it does not change any simulated value.

* **The omegas are variances, not standard deviations.** Table 4 labels them
  `omega^2 V` and `omega^2 CL`, and the reported confidence intervals settle it
  independently of the label. The 95% CI for `omega^2 V` is 0.073-0.23 about a
  point estimate of 0.15, a half-width of roughly 52%. For a **variance**
  estimated from 23 subjects the expected half-width is
  1.96 * sqrt(2/23) = 58%; for a **standard deviation** it would be
  1.96 * sqrt(1/(2*23)) = 29%. The observed 52% matches the variance scale and
  is far too wide for an SD. On the exponential IIV model this gives CVs of
  40.4% for V and 41.7% for CL.

* **Table 5 is not exactly reproducible from the published parameter set.**
  With the equations and Table 4 values encoded exactly - the machine-precision
  check above proves the encoding, so these are properties of the published
  numbers, not of this implementation - two systematic gaps remain.

  *Clearance runs high.* Back-solving Table 5's trough percentages implies a
  terminal half-life near 10 h in the 37-60 umol/L band; the published
  parameters give about 6 h, i.e. clearance roughly 1.4-fold too fast. The same
  1.4-fold factor appears against the study's **own observed data**: the
  cohort-median subject's trough works out to about 2.2 ug/mL against the
  observed median of 4.07 ug/mL (Table 2), which implies a clearance about
  0.70 times the published value. That two independent comparisons - the
  paper's simulation table and the paper's observed concentrations - both point
  to the same factor is what makes this a property of the published parameter
  set rather than a modelling choice here. The peak at the *end of the
  infusion* is unaffected (17.5 simulated against 17.45 observed), because it
  is governed by V, which is not in question; the peak at the paper's TDM
  sampling time 1 h later comes out 16% low (14.6), which is exactly the extra
  hour of too-fast decay.

  *The peak distribution is too wide.* `omega^2 V` = 0.15 gives a log-scale SD
  of 0.39 on peak concentration; solving Table 5's three peak bands for the
  lognormal that produces them gives about 0.25. The medians agree to within
  about 10%, so this is a spread mismatch, not a location one.

  Li 2025 does not report how creatinine was drawn inside each band, at which
  time the simulated peak and trough were read, or whether residual error was
  added to the simulated concentrations, so its Monte Carlo cannot be
  reproduced exactly in any case. No parameter was tuned; the discrepancy is
  reported instead. Users reproducing Li 2025's dosing recommendations from
  this model should expect the qualitative ordering to hold (it is asserted
  above) but not the percentages.

* **Publication year.** The DOI slug, the copyright line and the EuropePMC
  record all carry 2025 (accepted 18 September 2025); the journal assigned the
  article to volume 26, dated 2026. The model file is named `Li_2025_amikacin`
  after the DOI and copyright year, and the `reference` field records both.

## Assumptions in this vignette

* **Serum creatinine within a band is drawn uniformly.** Li 2025 states only
  that "Monte Carlo simulations were performed for 1000 random individuals with
  a random distribution of creatinine level within a quartiles range"; uniform
  is the natural reading of "random distribution ... within a range".

* **Peak is read at the end of the 0.5-h infusion and trough 30 min before the
  next dose.** These are the study's own TDM sampling times (Methods, "Blood
  sampling and concentration determination"). Reading the peak 1 h after the
  end of the infusion instead - the other rule the Methods gives - lowers the
  simulated peak by about 30% and moves it further from Table 5, so the
  end-of-infusion convention was used for the Table 5 comparison.

* **Weight, postmenstrual age and creatinine are held constant per subject
  over the simulated interval.** All three are time-varying in principle; the
  simulation covers at most 8 dosing intervals, over which the change is
  negligible.

* **Cohort sizes are 120 per arm** rather than the 1000 Li 2025 used, and only
  the postmenstrual-age / weight combination Table 5 reports in every
  creatinine band (PMA 31 weeks, WT 1.2 kg) is simulated, to keep the vignette
  inside the render budget. Monte Carlo noise on the reported percentages is
  roughly 4 percentage points, well below the systematic gaps discussed above.
  Table 5's peak percentages vary by under 2 points across its four PMA /
  weight combinations, because peak concentration is dose/V and V scales
  linearly with weight, so the omitted combinations add little.

* **No parameter value in this vignette or the model file came from anywhere
  other than the Li 2025 main text, its equations, or its Tables 1-5.** There
  is no supplement, and no author correspondence was needed.
