---
title: "Modakafusp alfa (Li 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modakafusp alfa (Li 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4.5, dpi = 96
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
set.seed(20250817)
```

# Model and source

Li 2025 reports two sequentially fitted models, extracted here as two model
files that share this article.

```{r models}
pk   <- rxode2::rxode(readModelDb("Li_2025_modakafuspAlfa"))
pkpd <- rxode2::rxode(readModelDb("Li_2025_modakafuspAlfa_mprotein"))
```

* Citation: `r pk$reference`
* Population PK model: `r pk$description`
* PK-PD model: `r pkpd$description`
* Article: <https://doi.org/10.1111/cts.70296>
* Supporting Information (open access, includes both NONMEM control streams and
  Figures S1-S12): <https://doi.org/10.1111/cts.70296>

Modakafusp alfa (TAK-573) is a first-in-class fusion protein: a humanized
anti-CD38 IgG4 kappa antibody with attenuated human interferon alfa-2b fused to
the C-terminus of each heavy chain. It was studied in the Phase 1/2 iinnovate-1
trial (NCT03215030) in relapsed or refractory multiple myeloma (RRMM).

# Population

```{r population}
pop <- pk$population
tibble::tibble(Field = names(pop),
               Value = vapply(pop, function(x) paste(
                 if (is.null(names(x))) as.character(x)
                 else paste0(names(x), " ", as.character(x)),
                 collapse = "; "), character(1))) |>
  knitr::kable()
```

The population PK analysis used 96 of the 100 enrolled RRMM patients (those with
at least one quantifiable serum concentration), contributing 2297 quantifiable
observations plus 806 below the 6.25 ng/mL limit of quantification, which were
handled as censored data by the Beal M3 likelihood method. The sequential PK-PD
analysis of serum M-protein used the 60 patients whose baseline M-protein was at
least 5 g/L, contributing 492 observations. Baseline demographics are Li 2025
Table 1: mean weight 79.9 kg (SD 23.0, range 60-169), mean age 63.1 years (SD
10.5, range 34-84), 42.7% female, 78.1% Caucasian, and mean serum albumin
3.72 g/dL (SD 0.58).

Immunogenicity was central to the PK model. 12.5% of PK-evaluable patients were
ADA-positive at baseline and 52.1% became ADA-positive on treatment, with titers
rising over repeated dosing (Table S2 and Figure S3).

# Source trace

Per-parameter provenance is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Li_2025_modakafuspAlfa.R` and
`inst/modeldb/specificDrugs/Li_2025_modakafuspAlfa_mprotein.R`. The table
collects them here.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` (CL) | 0.0553 L/day | Table 2 |
| `lvc` (Vc at 80.8 kg) | 5.08 L | Table 2 |
| `e_wt_vc` | 0.509 | Table 2 and footnote b |
| `lq` (Q) | 0.137 L/day | Table 2 |
| `lvp` (Vp) | 4.01 L | Table 2 |
| `lvmax` (Vmax) | 4.77 nmol/h | Table 2 |
| `lkm` (KM) | 1.26 nM | Table 2 |
| `lrtot_ada` (Rtot,ADA at titer 81) | 1.35 nM | Table 2 |
| `e_ada_titer_rtot_ada` | 0.412 | Table 2 (footnote c prints 4.12; see Errata) |
| `lkon_ada` | 3.6 1/(nM*h), fixed | Table 2, from Chen 2016 |
| `lkoff_ada_max` / `lkoff_ada_min` | 3600 / 0.036 1/h, fixed | Table 2, from Chen 2016 |
| `kdecay_ada` (kdec) | 0.0275 | Table 2 |
| `lkel_ada` | 0.210 1/h | Table 2 |
| PK IIV and residual error | omega^2 and sigma^2 | Table 2 |
| Molecular weight 186 kDa | 1 nM = 186 ng/mL | Supporting Information NONMEM streams (`DOSE/186.0`, `C*186`) |
| ADA gate at log3 titer > 3 | n/a | Supporting Information popPK stream (`L3OFFTITER = 3.0`) |
| `TVVc = 5.08*(WT/80.8)^0.509` | n/a | Table 2 footnote b |
| `TVRtot,ADA = 1.35*(Titer/81)^0.412` | n/a | Table 2 footnote c, corrected |
| `koff,ADA` titer function | n/a | Table 2 footnote d; Figure S6 |
| ADA binding ODEs (kon / koff / kel,ADA, Rfree = Rtot - complex) | n/a | Figure 1 and the popPK `$DES` block |
| `lrbase` (MP0 at 3.6 g/dL albumin) | 17.4 g/L | Table 3 |
| `e_alb_rbase` | -2.74 | Table 3 and footnote a |
| `lp` (kg) | 0.0303 1/week | Table 3 |
| `llambda` (kr) | 0.0568 1/week | Table 3 |
| `lkkillmax` | 0.816 1/week | Table 3 |
| `lec50` | 447 ng/mL | Table 3 |
| PD IIV, EC50 / kkill,max covariance | omega^2, -0.374 | Table 3 and footnote b |
| PD residual error | sigma^2 = 0.0161 (prop), 6.58 (add) | Table 3 |
| `dMP/dt = kg*MP - kkill,max*exp(-kr*t)*(Cp/(EC50+Cp))*MP` | n/a | Equation 1 |
| Objective response / progressive disease definitions | n/a | Methods section 2.5 |

# Structural verification of the ADA sub-model

The anti-drug-antibody layer is the distinctive part of this model, and it is
fully determined by the log3 ADA titer covariate. Both titer relationships can
therefore be checked exactly against the published typical-value curves, with no
simulation variability involved.

```{r ada-relationships}
# Both relationships are algebraic functions of the covariate alone, so a
# single dose-free observation record per titer level is enough to read them
# straight out of the compiled model.
ada_curve <- function(l3) {
  d <- data.frame(id = seq_along(l3), time = 0, amt = NA_real_, evid = 0L,
                  cmt = "central", WT = 80.8, ADA_TITER = l3)
  r <- rxode2::rxSolve(rxode2::zeroRe(pk), d, returnType = "data.frame",
                       addDosing = FALSE)
  if (is.null(r$id)) r$id <- 1L
  data.frame(l3titer = l3[r$id], rtot_ada = r$rtot_ada,
             koff_per_h = r$koff_ada / 24)
}
ada <- ada_curve(0:13)
knitr::kable(ada, digits = c(0, 3, 2),
             col.names = c("log3(ADA titer)", "Rtot,ADA (nM)", "koff,ADA (1/h)"))
```

Li 2025 Figure S5 plots the typical `Rtot,ADA` against log3 titer: it starts at
1.35 nM at a log3 titer of 4 (a reciprocal titer of 81, the reference), passes
through roughly 12-13 nM at 9, and reaches about 79 nM at 13. Figure S6 plots
`koff,ADA` starting at 3.6e3 1/h at a log3 titer of 4 and decaying toward the
3.6e-2 1/h asymptote. Both are reproduced exactly.

```{r ada-assertions}
tol <- function(x, target, pct) abs(x / target - 1) <= pct / 100
stopifnot(
  # ADA-negative (and any titer at or below the log3 gate of 3) gives no pool
  ada$rtot_ada[ada$l3titer == 0] == 0,
  ada$rtot_ada[ada$l3titer == 3] == 0,
  # Table 2 reference value
  tol(ada$rtot_ada[ada$l3titer == 4], 1.35, 0.5),
  # Figure S5 typical-value line
  tol(ada$rtot_ada[ada$l3titer == 9], 12.98, 2),
  tol(ada$rtot_ada[ada$l3titer == 13], 79.3, 2),
  # Table 2 koff,ADA,max and the Figure S6 value at the top of the observed range
  tol(ada$koff_per_h[ada$l3titer == 4], 3600, 0.5),
  tol(ada$koff_per_h[ada$l3titer == 13], 288, 2)
)
"ADA titer relationships reproduce Table 2, Figure S5 and Figure S6"
```

```{r ada-figure, fig.cap = "Reproduction of Li 2025 Figures S5 and S6: the typical total ADA pool and the drug-ADA dissociation rate constant as functions of the log3 ADA titer, over the titer range observed in the study."}
ada |>
  filter(l3titer >= 4) |>
  tidyr::pivot_longer(c(rtot_ada, koff_per_h),
                      names_to = "quantity", values_to = "value") |>
  mutate(quantity = recode(quantity,
                           rtot_ada   = "Rtot,ADA (nM) -- Figure S5",
                           koff_per_h = "koff,ADA (1/h) -- Figure S6")) |>
  ggplot(aes(l3titer, value)) +
  geom_line(colour = "purple", linewidth = 1.1) +
  geom_point(colour = "purple", size = 1.4) +
  scale_y_log10() +
  facet_wrap(~quantity, scales = "free_y") +
  labs(x = expression(log[3]*"(ADA titer)"), y = NULL) +
  theme_bw()
```

# Typical-value pharmacokinetics

Doses are converted from mg to nmol with the model's 186 kDa molecular weight
(`nmol = mg / 0.186`). Infusions are the 1-h infusion used for the published
simulations.

```{r helpers}
MW_KDA   <- 186
MG_TO_NMOL <- 1000 / MW_KDA        # 1 mg = 1000/186 nmol for a 186 kDa protein

# Build an event data frame. Observation rows use cmt = "central", an ODE state
# name, never an algebraic observable name.
build_events <- function(id, amt_nmol, wt, alb = NA_real_, l3titer = 0,
                         dose_times = 0, obs_times) {
  dose <- data.frame(
    id = rep(id, each = length(dose_times)),
    time = rep(dose_times, times = length(id)),
    amt = rep(amt_nmol, each = length(dose_times)),
    dur = 1 / 24, evid = 1L, cmt = "central"
  )
  obs <- data.frame(
    id = rep(id, each = length(obs_times)),
    time = rep(obs_times, times = length(id)),
    amt = NA_real_, dur = NA_real_, evid = 0L, cmt = "central"
  )
  ev <- rbind(dose, obs)
  ev <- ev[order(ev$id, ev$time, -ev$evid), ]
  idx <- match(ev$id, id)
  ev$WT <- wt[idx]
  ev$ALB <- alb[idx]
  ev$ADA_TITER <- l3titer
  ev
}

solve_pop <- function(mod, ev, typical = FALSE) {
  m <- if (typical) rxode2::zeroRe(mod) else mod
  out <- rxode2::rxSolve(m, ev, omega = if (typical) NULL else mod$omega,
                         returnType = "data.frame", addDosing = FALSE,
                         atol = 1e-10, rtol = 1e-8)
  if (is.null(out$id)) out$id <- 1L   # rxSolve drops id for a single subject
  out
}
```

```{r typical-pk, fig.cap = "Typical-value serum modakafusp alfa after a single 1-h IV infusion of 120 mg or 240 mg to an 80.8 kg ADA-negative patient. The near-flat early phase followed by a steep decline is the signature of the saturable Michaelis-Menten elimination pathway."}
tgrid <- c(seq(0, 2, by = 1 / 48), seq(2.05, 28, by = 0.05))
typ <- lapply(c(120, 240), function(mg) {
  ev <- build_events(1L, mg * MG_TO_NMOL, 80.8, obs_times = tgrid)
  solve_pop(pk, ev, typical = TRUE) |> mutate(dose = paste(mg, "mg"))
}) |> bind_rows()

ggplot(typ, aes(time, pmax(Cc, 0.1), colour = dose)) +
  geom_line(linewidth = 0.9) +
  scale_y_log10() +
  geom_hline(yintercept = 6.25, linetype = "dashed", colour = "grey40") +
  labs(x = "Time (day)", y = "Serum modakafusp alfa (ng/mL)",
       colour = "Dose",
       caption = "Dashed line: 6.25 ng/mL assay LLOQ") +
  theme_bw()
```

```{r typical-cmax}
typ_cmax <- typ |> group_by(dose) |> summarise(cmax = max(Cc), .groups = "drop")
knitr::kable(typ_cmax, digits = 0,
             col.names = c("Dose", "Typical Cmax (ng/mL)"))
# Cmax for an IV bolus / short infusion is dose / Vc, independent of the
# molecular weight, so this is a direct check on the Vc parameterisation.
# Look the value up by key rather than by row position.
expected <- c(`120 mg` = 120, `240 mg` = 240) * 1000 / 5.08
stopifnot(setequal(typ_cmax$dose, names(expected)))
stopifnot(all(abs(typ_cmax$cmax / expected[typ_cmax$dose] - 1) < 0.02))
"Typical Cmax equals dose / Vc to within 2%"
```

# Virtual cohort

Li 2025 generated 2500 virtual RRMM patients by tilted bootstrap so that mean
body weight and albumin matched iinnovate-1. The weight quartile cut-points used
for Figure 3 (`<= 62`, `(62, 77]`, `(77, 96]`, `> 96` kg) pin that distribution
down: a log-normal with a median of 77 kg and a log-scale SD of 0.322 reproduces
all three quartiles and gives a mean of 81 kg, close to the reported 79.9 kg.
Albumin is drawn to the Table 1 mean and SD, truncated at the reported range.

Cohorts here are **200 subjects per arm**, the nlmixr2lib vignette cap, rather
than the paper's 2500 (Figure 3) or 1000 (Figure S12).

```{r cohort}
n_arm <- 200L
cohort <- tibble::tibble(
  id  = seq_len(n_arm),
  WT  = pmin(pmax(exp(rnorm(n_arm, log(77), 0.322)), 40), 175),
  ALB = pmin(pmax(rnorm(n_arm, 37.5, 5.7), 19), 46)   # g/L (Table 1: 3.75 g/dL)
)
cohort |>
  summarise(`Median WT (kg)` = median(WT), `Q1 WT` = quantile(WT, .25),
            `Q3 WT` = quantile(WT, .75), `Mean WT` = mean(WT),
            `Mean ALB (g/L)` = mean(ALB)) |>
  knitr::kable(digits = 1)
```

# Weight-based versus fixed dosing (replicates Figure 3)

The paper's headline PK result is that 120 mg and 240 mg fixed doses give
exposure comparable to 1.5 and 3.0 mg/kg, which supported the switch to fixed
dosing. Figure 3 simulates a single 28-day cycle, during which almost all
patients are still ADA-negative, so `ADA_TITER` is 0 throughout.

```{r fig3-sim}
arms <- tibble::tribble(
  ~treatment,    ~mgkg, ~mg,
  "1.5 mg/kg",     1.5,  NA,
  "120 mg",         NA, 120,
  "3 mg/kg",       3.0,  NA,
  "240 mg",         NA, 240
)
nca_grid <- c(seq(0, 2, by = 1 / 24), seq(2 + 1 / 12, 28, by = 1 / 12))

# rxSolve already returns the covariate columns it was given, so weight is read
# back from the cohort by id rather than joined (a join would collide with the
# WT column already present and silently rename it).
wt_band <- function(wt) {
  cut(wt, c(-Inf, 62, 77, 96, Inf),
      labels = c("Q1: <= 62", "Q2: (62, 77]", "Q3: (77, 96]", "Q4: > 96"))
}

sim3 <- lapply(seq_len(nrow(arms)), function(i) {
  a <- arms[i, ]
  amt <- if (is.na(a$mg)) a$mgkg * cohort$WT * MG_TO_NMOL
         else rep(a$mg * MG_TO_NMOL, n_arm)
  ev <- build_events(cohort$id, amt, cohort$WT, obs_times = nca_grid)
  solve_pop(pk, ev) |>
    mutate(treatment = a$treatment, amt_nmol = amt[match(id, cohort$id)])
}) |> bind_rows() |>
  mutate(wt_kg = cohort$WT[match(id, cohort$id)],
         wt_quartile = wt_band(wt_kg))
stopifnot(!anyNA(sim3$wt_kg))
stopifnot(dplyr::n_distinct(sim3$id) == n_arm,
          nrow(sim3) == 4L * n_arm * length(nca_grid))
```

## NCA with PKNCA

```{r pknca}
conc_df <- sim3 |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment, wt_quartile)

dose_df <- sim3 |>
  group_by(treatment, id) |>
  summarise(time = 0, amt = first(amt_nmol), .groups = "drop") |>
  mutate(wt_quartile = wt_band(cohort$WT[match(id, cohort$id)])) |>
  select(id, time, amt, treatment, wt_quartile)

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | treatment + id,
                             concu = "ng/mL", timeu = "day")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "nmol")
intervals <- data.frame(start = 0, end = 28,
                        cmax = TRUE, tmax = TRUE, auclast = TRUE)
nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                      intervals = intervals))
nca_ind <- as.data.frame(nca) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "auclast")) |>
  mutate(PPORRES = ifelse(PPTESTCD == "auclast", PPORRES * 24, PPORRES))
# auclast is ng*day/mL from a day time base; x24 puts it on the paper's
# ng*h/mL axis.
```

```{r fig3-plot, fig.cap = "Replicates Li 2025 Figure 3: model-predicted Cmax and AUC0-tau after a single cycle, overall and by body-weight quartile, comparing weight-based with fixed dosing."}
plt <- nca_ind |>
  filter(PPTESTCD %in% c("cmax", "auclast")) |>
  left_join(dose_df |> select(id, treatment, wt_quartile),
            by = c("id", "treatment")) |>
  mutate(
    pair = ifelse(treatment %in% c("1.5 mg/kg", "120 mg"),
                  "1.5 mg/kg vs 120 mg", "3 mg/kg vs 240 mg"),
    mode = ifelse(grepl("kg", treatment), "Weight-based", "Fixed"),
    param = ifelse(PPTESTCD == "cmax", "Cmax (ng/mL)",
                   "AUC0-tau (ng*h/mL)"),
    wt_quartile = as.character(wt_quartile)
  )
plt_all <- bind_rows(plt, mutate(plt, wt_quartile = "Overall")) |>
  mutate(wt_quartile = factor(wt_quartile,
                              levels = c("Overall", "Q1: <= 62", "Q2: (62, 77]",
                                         "Q3: (77, 96]", "Q4: > 96")))
ggplot(plt_all, aes(wt_quartile, PPORRES, fill = mode)) +
  geom_boxplot(outlier.shape = NA, position = position_dodge(0.8)) +
  stat_summary(fun = mean, geom = "point", colour = "red", size = 1.4,
               position = position_dodge(0.8)) +
  facet_grid(param ~ pair, scales = "free_y") +
  coord_cartesian(ylim = c(0, NA)) +
  labs(x = "Body weight (kg)", y = NULL, fill = "Dosing") +
  theme_bw() + theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

The qualitative conclusion reproduces: within each pair, the fixed and
weight-based distributions overlap closely overall, the weight-based arm's
exposure is flatter across weight quartiles, and the fixed arm trends upward in
the lightest quartile and downward in the heaviest. That is the pattern Li 2025
used to argue that fixed dosing is acceptable when the weight exponents on CL
and Vc are both below about 0.5.

## Comparison against the published simulations

Li 2025 reports Figure 3 only as boxplots, so the reference values below were
read off the published figure and are accurate to roughly 10%.

```{r fig3-comparison}
reference <- tibble::tribble(
  ~treatment,   ~cmax,  ~auclast,
  "1.5 mg/kg",  22000,  0.88e6,
  "120 mg",     23500,  0.90e6,
  "3 mg/kg",    44000,  2.30e6,
  "240 mg",     48000,  2.40e6
)
sim_for_cmp <- nca_ind |>
  filter(PPTESTCD %in% c("cmax", "auclast")) |>
  select(treatment, PPTESTCD, PPORRES)

cmp <- nlmixr2lib::ncaComparisonTable(
  sim_for_cmp, reference, by = "treatment",
  units = c(cmax = "ng/mL", auclast = "ng*h/mL")
)
knitr::kable(cmp)
attr(cmp, "footnote")
```

`Cmax` reproduces the published medians closely in all four arms, within 10%.
`AUClast` does not: the simulated medians run about 1.2- to 1.5-fold high at the
120 mg / 1.5 mg/kg dose level and about 2.1- to 2.2-fold high at 240 mg /
3 mg/kg, and the discrepancy grows with dose.

That split is itself diagnostic, and it points at a specific gap in what the
paper publishes rather than at the structural model. `Cmax` after an IV
infusion is dose divided by `Vc`, so it depends on exactly one random effect
(`etalvc`), whose variance Table 2 does report. `AUC` under parallel linear and
saturable elimination depends jointly on `CL`, `Vc`, `Q`, `Vp` and `Vmax`. Those
five etas were estimated in the source as a full `$OMEGA BLOCK(5)`, but Table 2
publishes only the five diagonal variances; the ten covariances are not reported
anywhere in the article or the Supporting Information. Carrying the etas as
independent (the only option that does not invent numbers) inflates the spread
of the exposure distribution, which is visible directly:

```{r spread-check}
sim3 |>
  filter(treatment == "1.5 mg/kg", abs(time - 7) < 1e-6) |>
  summarise(`5th pct` = quantile(Cc, 0.05), Median = median(Cc),
            `95th pct` = quantile(Cc, 0.95)) |>
  knitr::kable(digits = 3)
```

Li 2025 Figure S12 panel A shows a 90% prediction interval about three orders of
magnitude wide at one week after a 1.5 mg/kg dose; the interval above is
substantially wider. Because AUC is a convex functional of the disposition
parameters, over-dispersion inflates the mean more than the median and skews the
distribution, exactly as observed (the simulated mean-to-median ratio is larger
than the published figure's). No parameter was adjusted to close the gap.

# Serum M-protein response (replicates Figure S12)

Figure S12 is the cleanest end-to-end validation target in the paper because the
immunogenicity scenario is fully specified: 1.5 mg/kg Q4W for 10 cycles at a
**constant** log3 ADA titer of 0 (ADA-negative), 9, or 13, rather than the
bootstrapped individual titer trajectories used for Figure 4.

Objective response and progressive disease follow the Methods section 2.5
definitions, and dropout for progressive disease is applied as the paper
describes.

```{r pkpd-sim}
cycle_days <- seq(0, 9 * 28, by = 28)
obs_days   <- seq(0, 280, by = 7)

pkpd_run <- function(l3) {
  ev <- build_events(cohort$id, 1.5 * cohort$WT * MG_TO_NMOL,
                     cohort$WT, cohort$ALB, l3titer = l3,
                     dose_times = cycle_days, obs_times = obs_days)
  solve_pop(pkpd, ev) |>
    arrange(id, time) |>
    group_by(id) |>
    mutate(
      mp_base = first(mprotein),
      nadir   = cummin(mprotein),
      pchg    = 100 * (mprotein / mp_base - 1),
      # Li 2025 section 2.5: PD is a >= 25% increase relative to the lowest
      # value with an absolute increase of >= 5 g/L, or an absolute increase
      # of >= 10 g/L when baseline is >= 50 g/L.
      pd      = time > 0 &
        ((mprotein >= 1.25 * nadir & mprotein - nadir >= 5) |
           (mp_base >= 50 & mprotein - mp_base >= 10)),
      onstudy = cumsum(pd) == 0
    ) |>
    ungroup() |>
    mutate(l3titer = l3)
}
pkpd_sim <- lapply(c(0, 9, 13), pkpd_run) |> bind_rows()
stopifnot(dplyr::n_distinct(pkpd_sim$id) == n_arm,
          all(c(0, 9, 13) %in% pkpd_sim$l3titer))
```

```{r pkpd-mprotein, fig.cap = "Replicates Li 2025 Figure S12 panel B: median serum M-protein change from baseline among patients still on study, after 1.5 mg/kg Q4W for 10 cycles at a constant log3 ADA titer of 0, 9 or 13."}
mp_med <- pkpd_sim |>
  filter(onstudy) |>
  group_by(l3titer, time) |>
  summarise(median = median(pchg),
            lo = quantile(pchg, 0.05), hi = quantile(pchg, 0.95),
            .groups = "drop") |>
  mutate(titer = factor(l3titer))

ggplot(mp_med, aes(time / 7, median, colour = titer, fill = titer)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.12, colour = NA) +
  geom_line(linewidth = 1) +
  labs(x = "Time since first dose (week)",
       y = "Serum M-protein (% change from baseline)",
       colour = expression(log[3]*"(ADA titer)"),
       fill = expression(log[3]*"(ADA titer)")) +
  theme_bw()
```

```{r pkpd-orr, fig.cap = "Replicates Li 2025 Figure S12 panel C: objective response rate at the end of each dosing cycle. Objective response is a decrease of at least 50% in serum M-protein from baseline."}
orr <- pkpd_sim |>
  filter(time %in% (seq_len(10) * 28)) |>
  group_by(l3titer, cycle = time / 28) |>
  summarise(responders = sum(onstudy & pchg <= -50),
            remaining  = sum(onstudy),
            orr_itt    = 100 * responders / n_arm,
            .groups = "drop") |>
  mutate(titer = factor(l3titer))
stopifnot(nrow(orr) == 30L, all(orr$remaining > 0))

ggplot(orr, aes(cycle, orr_itt, colour = titer)) +
  geom_line(linewidth = 1) + geom_point() +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Dosing cycle", y = "Objective response rate (%)",
       colour = expression(log[3]*"(ADA titer)")) +
  theme_bw()
```

```{r pkpd-table}
summary_tbl <- orr |>
  group_by(titer) |>
  summarise(`Peak ORR (%)` = max(orr_itt),
            `Retention cycle 1 -> 10 (%)` =
              100 * remaining[cycle == 10] / remaining[cycle == 1],
            .groups = "drop") |>
  left_join(
    mp_med |> filter(time == 280) |>
      transmute(titer, `Median M-protein change at week 40 (%)` = median),
    by = "titer"
  ) |>
  mutate(
    `Published peak ORR (%)`  = c(40, 28, 12),
    `Published retention (%)` = c(58, 52, 30),
    `Published week-40 change (%)` = c(-86, -73, -43)
  )
knitr::kable(summary_tbl, digits = 0)
```

All three published quantities reproduce within about ten percentage points, and
the **ordering and spacing of the ADA effect** is right: response is nearly
halved between a log3 titer of 9 and 13, which is the paper's point that
high-titer immunogenicity materially erodes efficacy. The **dropout dynamics**
track Figure S12 panel C, which is a joint check on the M-protein growth rate,
the resistance rate, and the progressive-disease rule. The **week-40 median
M-protein change** at a log3 titer of 9 is essentially exact.

The residual bias is in the same direction as the AUC discrepancy above and has
the same cause: the over-dispersed exposure distribution that follows from the
unpublished `$OMEGA` covariances feeds directly into the saturable kill term, so
objective response rate and retention both run a little high and the week-40
medians a little more negative than published. The effect is far smaller here
than on AUC because the drug effect saturates: `EC50` is 447 ng/mL, so the kill
term is already close to its maximum whenever concentrations are high, and
response is therefore much less than proportional to exposure. A two-fold error
in AUC buys only a modest error in response.

```{r pkpd-assertions}
stopifnot(
  # The ADA gradient is monotone in the published direction
  all(diff(summary_tbl$`Peak ORR (%)`) < 0),
  all(diff(summary_tbl$`Median M-protein change at week 40 (%)`) > 0),
  # All three published quantities reproduce within 12 percentage points
  all(abs(summary_tbl$`Peak ORR (%)` -
            summary_tbl$`Published peak ORR (%)`) < 12),
  all(abs(summary_tbl$`Retention cycle 1 -> 10 (%)` -
            summary_tbl$`Published retention (%)`) < 12),
  all(abs(summary_tbl$`Median M-protein change at week 40 (%)` -
            summary_tbl$`Published week-40 change (%)`) < 12)
)
"ORR, retention and M-protein response reproduce Figure S12 within 12 points"
```

# Assumptions and deviations

## Errata and corrections to the source

* **Table 2 footnote c prints the wrong exponent.** The footnote reads
  `TVRtot,ADA (nM) = 1.35 * (Titer/81)^4.12`, but Table 2's own parameter row
  gives the `Rtot,ADA`-titer power as **0.412** with a 4.29% RSE. The model uses
  0.412. Three independent lines of evidence settle it: the typical-value line in
  Figure S5 runs from 1.35 nM at a log3 titer of 4 to about 79 nM at 13, which is
  `1.35 * (3^9)^0.412 = 79.3` and is many orders of magnitude away from what 4.12
  would give; observed reciprocal titers exceeded 164 000 (Figure S4 legend), at
  which an exponent of 4.12 would imply a physically impossible ADA
  concentration; and the NONMEM control stream in the Supporting Information
  hardcodes `**0.412`.
* **The PK-PD control stream carries a stale `kdec`.** The PK-PD stream sets
  `KD_TITERDEC = 0.210`, which is the value of `kel,ADA`, not of `kdec`. Table 2
  reports `kdec = 0.0275` with a 5.23% RSE, the popPK stream estimates it as
  `THETA(13)`, and Figure S6 is only consistent with 0.0275 (at 0.210 the curve
  would already be at the minimum by a log3 titer of 13, whereas the figure shows
  about 3.4e2 1/h there). The model uses 0.0275 in both files.
* **Table 1's M-protein column counts do not sum to 60.** The sex, ECOG and race
  counts in the "Serum M-protein analysis population" column sum to 99, while the
  Results text states that 60 patients formed the M-protein evaluable population.
  The model records `n_subjects = 60` per the Results text and reproduces the
  Table 1 percentages as published.
* **`koff,ADA,max` is labelled "when ADA titer = 0".** The published function
  evaluates to 3600 1/h at a log3 titer of 4, not at 0, and exceeds 3600 below
  that. This is immaterial: `Rtot,ADA` is gated to zero at or below a log3 titer
  of 3, so no ADA binding occurs in that range and `koff,ADA` never acts.

## Encoding decisions

* **Time base.** The source NONMEM streams work in hours; these model files work
  in days, which suits the 21- and 28-day cycles. Per-hour rate constants from
  Table 2 are multiplied by 24 and the per-week PD rate constants from Table 3
  are divided by 7, mirroring the source stream's own `/24/7` conversion.
* **Molecular weight.** 186 kDa, taken from the control streams
  (`IN = DOSE/186.0` and `IPRED = (A(1)/S1)*186.0`); the article text does not
  state it. Doses must be supplied in nmol.
* **The `complex` state is a concentration, not an amount.** `A(3)` in the source
  `$DES` is written in nM, with the return flux to the central compartment
  multiplied back up by `Vc`. That is replicated rather than rescaled, because
  with a time-varying body weight (and hence a time-varying `Vc`) the two
  formulations are not identical.
* **Free ADA is `Rtot,ADA` minus the complex.** Li 2025 states that "Rtot,ADA was
  assumed to be constant at a given time-varying titer level (i.e., no turnover
  of ADA)", so the pool is implicitly replenished as complex is cleared.
* **The ADA gate at a log3 titer above 3** is taken from the control stream
  (`L3OFFTITER = 3.0`); the article text says only that the pool is zero during
  ADA negativity.
* **Albumin is in g/L.** The nlmixr2lib canonical unit for `ALB` is g/L, whereas
  Li 2025 reports g/dL and centres the M-protein baseline effect on 3.6 g/dL. The
  model centres on 36 g/L; because the covariate enters as a ratio, the effect is
  unchanged.

## Known limitations

* **The `$OMEGA BLOCK(5)` covariances are not published.** Table 2 reports the
  five diagonal variances of the correlated CL / Vc / Q / Vp / Vmax block but
  none of the ten covariances, and they appear nowhere in the Supporting
  Information (the control stream lists only initial estimates). The five etas
  are therefore carried as independent, which is the only encoding that does not
  invent numbers. The consequence is quantified above: typical-value predictions
  and single-eta statistics such as `Cmax` reproduce, while joint-eta statistics
  such as `AUC0-tau`, the width of the prediction interval, and the objective
  response rate are over-dispersed and biased high. Anyone using these models for
  population simulation should treat the exposure spread as an upper bound.
* **Sequential fitting.** The source fitted the PD with each patient's post hoc
  PK parameters held fixed. `Li_2025_modakafuspAlfa_mprotein` carries the
  published typical PK values and their IIV so that it can be simulated on its
  own; that composition is not identical to the sequential estimation the authors
  performed.
* **Cohort size.** 200 subjects per arm here versus 2500 (Figure 3) and 1000
  (Figure S12) in the paper.
* **The virtual cohort is reconstructed, not the paper's.** Li 2025 used a tilted
  bootstrap over a correlated virtual RRMM population; weight and albumin here
  are drawn independently from marginal distributions matched to Table 1 and to
  the Figure 3 weight quartile cut-points.
* **Figure 4 is not reproduced.** It requires bootstrapped individual longitudinal
  ADA titer trajectories from the 1.5 and 3.0 mg/kg Q4W cohorts, which are shown
  only graphically in Figure S3. Figure S12, whose constant-titer scenarios are
  fully specified, is reproduced instead.
* **Reference values read from figures.** The Figure 3 and Figure S12 comparison
  values were digitised from the published figures and are accurate to roughly
  10%; the paper reports no corresponding numeric table.
