---
title: "TQ-B3203 (Li 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TQ-B3203 (Li 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r load-model}
mod <- readModelDb("Li_2023_TQB3203")
ui <- rxode2::rxode(mod)
```

# Model and source

* Citation: `r ui$reference`
* Article: <https://doi.org/10.3389/fphar.2023.1102244>
* Free full text: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9885713/>

TQ-B3203 is a semisynthetic camptothecin (topoisomerase I inhibitor) derivative
carrying an aliphatic chain, formulated as a liposome injection (TLI) to reduce
toxicity and prolong circulation. Li 2023 is the first population PK analysis of
the compound, built from the phase I dose-escalation trial NCT03447145.

A **three-compartment model with first-order elimination** described the data
best, parameterized as central volume `V1`, shallow peripheral volume `V2`, deep
peripheral volume `V3`, elimination clearance `CL`, and inter-compartmental
clearances `CL2` (central to shallow) and `CL3` (central to deep). Only the
parent compound was modelled; conversion to SN-38 was under 5% (Discussion).

# Population

`r ui$population$n_observations` plasma concentrations from
`r ui$population$n_subjects` Chinese patients with advanced solid tumors over
`r ui$population$n_episodes` treatment episodes (Results 3.1). Patients were
`r ui$population$age_range` with weight `r ui$population$weight_range`, BMI
`r ui$population$bmi_range`, and lean body mass `r ui$population$lbm_range`;
`r ui$population$sex_female_pct`% were female (Table 1). Doses were
`r ui$population$dose_range`

Sampling was pre-dose, 45 and 90 min after infusion start, then 0.5, 1, 2, 4, 8,
12, 24, 48, 72 and 96 h after the **end** of the infusion, so the last sample is
97.5 h after the start of dosing. That grid is reproduced exactly below wherever
the comparison is against a published NCA number.

The eligibility criteria restricted BMI to 18.5-26 kg/m^2 and required normal
primary organ function, so the covariate ranges are narrow. The authors
explicitly caution that extrapolation beyond them is not supported.

The same information is available programmatically via
`rxode2::rxode(readModelDb("Li_2023_TQB3203"))$population`.

# Source trace

Every `ini()` entry in `inst/modeldb/specificDrugs/Li_2023_TQB3203.R` carries an
in-file comment naming its origin. Collected here for review:

| Equation / parameter | Value | Source location |
|---|---|---|
| Structural model (3-cmt, first-order elimination) | n/a | Results 3.3, first sentence |
| `lvc` (V1) | 4.81 L | Table 4; Eq. 9; Results 3.3; Discussion |
| `lvp` (V2) | 24.44 L | Table 4 |
| `lvp2` (V3) | 27.98 L | Table 4 |
| `lcl` (CL) | 3.97 L/h | Table 4; Eq. 10; Abstract; Discussion |
| `lq` (CL2) | 1.95 L/h | Table 4 |
| `lq2` (CL3) | 10.58 L/h | Table 4 |
| `e_bmi_cl` | 0.78 | Table 4 "BMI on CL"; exponent in Eq. 10; Results 3.3 |
| `e_dbil_cl` | -0.24 | Table 4 "DBIL on CL"; exponent in Eq. 10; Results 3.3 |
| `e_dbil_q` | -1.77 | Table 4 "DBIL on CL2" |
| `e_lbm_q` | -2.55 | Table 4 "LBW on CL2" |
| `e_lbm_vc` | 1.18 | Table 4 "LBW on V1"; exponent in Eq. 9; Results 3.3 |
| `e_lbm_vp` | -1.41 | Table 4 "LBW on V2" |
| Covariate functional form `(Cov/Cov_median)^theta` | n/a | Eq. 7 |
| BMI median 23.44 kg/m^2 | normalizing constant | Results 3.3; Table 1 |
| DBIL median 2.82 umol/L | normalizing constant | Results 3.3; Table 1 |
| LBM (LBW) median 49.53 kg | normalizing constant | Results 3.3; Table 1 |
| `etalcl` (omega^2 CL) | 0.043 | Table 4 IIV section |
| `etalvp2` (omega^2 V3) | 0.117 | Table 4 IIV section |
| `etalq` (omega^2 CL2) | 0.573 | Table 4 IIV section |
| `etalq2` (omega^2 CL3) | 0.290 | Table 4 IIV section |
| `etalvc` | fixed(0) | Eq. 9 prints exp(eta_V1); Table 4 reports no omega^2 for V1 |
| `expSd` (log-additive sigma) | 0.200 | Table 4 "Residual variability (sigma) stdev0"; Results 3.3 |
| Exponential IIV model | n/a | Eq. 1 |
| Log-additive residual model | n/a | Results 3.3 |
| No IOV | n/a | Results 3.3 (IOV tested on CL and V1, rejected) |
| Which covariates are in the final model | n/a | Table 3 step 9 |

```{r show-model}
ui
```

# Validation strategy

Li 2023 publishes three kinds of number this model can be checked against:

1. **Closed-form identities.** For a linear model, AUC to infinity is exactly
   `Dose / CL`, and `CL` depends only on BMI and DBIL. Those give gates with no
   free parameters and no sampling noise.
2. **Figure 9**, a sensitivity analysis printing the percentage change in
   AUC0-t and Cmax when one covariate is moved to its 5th or 95th percentile.
3. **Table 2**, the observed non-compartmental analysis by dose level.

Sections below take them in that order.

## Closed-form gate 1: AUC to infinity equals Dose / CL

`CL2` and `CL3` redistribute drug but eliminate none of it, so they cannot
change total exposure; `V1`, `V2`, `V3` change only the shape. Therefore
AUC0-inf must equal `Dose / CL` for every subject, whatever their covariates
and random effects. Both sides here use the *same* drawn parameters, so the
only difference is numerical integration error and a tight bound is correct.

```{r gate-aucinf}
# Median covariate values from Table 1; dose 25 mg/m^2 as used for Figure 9.
med <- list(BMI = 23.44, DBIL = 2.82, LBM = 49.53, BSA = 1.678)
dose_typ <- 25 * med$BSA

# Dense grid, long enough to resolve the terminal phase without decaying into
# solver noise (terminal half-life is ~16 h, so 336 h is ~21 half-lives).
ev_dense <- rxode2::et(amt = dose_typ, dur = 1.5, cmt = "central") |>
  rxode2::et(seq(0, 336, by = 0.02))

sim_dense <- rxode2::rxSolve(
  mod, ev_dense,
  params = c(BMI = med$BMI, DBIL = med$DBIL, LBM = med$LBM),
  omega = NA, returnType = "data.frame"
)

trapz <- function(t, y) sum(diff(t) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)

auc_solved <- trapz(sim_dense$time, sim_dense$Cc)
cl_typ <- unique(round(sim_dense$cl, 10))
auc_closed <- 1000 * dose_typ / cl_typ   # 1000 converts mg/L to ng/mL

stopifnot(
  length(cl_typ) == 1L,
  all(sim_dense$Cc >= 0),
  # Pure numerical-integration error: the two sides share every parameter.
  abs(auc_solved / auc_closed - 1) < 0.005
)

data.frame(
  Quantity = c("CL (L/h)", "AUC0-inf solved (ng*h/mL)",
               "AUC0-inf = 1000*Dose/CL (ng*h/mL)", "Relative difference (%)"),
  Value = c(cl_typ, auc_solved, auc_closed,
            100 * (auc_solved / auc_closed - 1))
) |>
  knitr::kable(digits = 4, caption = "Closed-form AUC identity, typical subject at 25 mg/m^2.")
```

## Closed-form gate 2: Cmax occurs at the end of the infusion

Results 3.2 states that "Cmax always appears at the end of intravenous
injection (1.5 h)". For a linear model with first-order elimination and a
zero-order input, the peak is at end of infusion exactly.

```{r gate-tmax}
tmax_solved <- sim_dense$time[which.max(sim_dense$Cc)]
stopifnot(abs(tmax_solved - 1.5) < 1e-8)
cat(sprintf("Tmax = %.3f h (paper: 1.5 h, Results 3.2 and Table 2)\n", tmax_solved))
```

## Closed-form gate 3: the covariate exponents on clearance

Because AUC0-inf is `Dose / CL` and `CL = 3.97 * (BMI/23.44)^0.78 *
(DBIL/2.82)^-0.24`, moving one covariate must change AUC0-inf by exactly the
reciprocal power. Moving LBM must not change AUC0-inf **at all**, because LBM
touches only `V1`, `V2` and `CL2` -- none of which eliminate drug. That null is
the sharpest single check in this vignette: it fails if any LBM exponent has
been mis-attached to a clearance term.

```{r gate-cov-exponents}
cl_at <- function(BMI = med$BMI, DBIL = med$DBIL, LBM = med$LBM) {
  s <- rxode2::rxSolve(
    mod, rxode2::et(amt = dose_typ, dur = 1.5, cmt = "central") |> rxode2::et(0:2),
    params = c(BMI = BMI, DBIL = DBIL, LBM = LBM),
    omega = NA, returnType = "data.frame"
  )
  unique(round(s$cl, 10))
}

cov_checks <- tibble::tribble(
  ~Covariate, ~Value,  ~Expected,
  "BMI",      18.64,   (18.64 / 23.44)^0.78,
  "BMI",      28.97,   (28.97 / 23.44)^0.78,
  "DBIL",     1.50,    (1.50 / 2.82)^-0.24,
  "DBIL",     7.90,    (7.90 / 2.82)^-0.24,
  "LBM",      32.09,   1,
  "LBM",      59.40,   1
) |>
  rowwise() |>
  mutate(
    Observed = dplyr::case_when(
      Covariate == "BMI"  ~ cl_at(BMI = Value),
      Covariate == "DBIL" ~ cl_at(DBIL = Value),
      Covariate == "LBM"  ~ cl_at(LBM = Value)
    ) / cl_typ
  ) |>
  ungroup() |>
  mutate(`% diff` = 100 * (Observed / Expected - 1))

# Exact algebraic identities, not a simulated comparison.
stopifnot(max(abs(cov_checks$`% diff`)) < 1e-6)

cov_checks |>
  rename("CL ratio expected" = Expected, "CL ratio observed" = Observed) |>
  knitr::kable(digits = 6, caption = "CL covariate exponents against the printed power model (Eq. 10 and Table 4).")
```

# Replicating Figure 9 (covariate sensitivity)

Figure 9 prints, inside the plot panels, the percentage change in AUC0-t and
Cmax when one covariate is set to its 5th or 95th percentile with the others
held at their medians. The two bar labels for each covariate sum to the total
quoted in Results 3.5, which confirms the reading:

| Covariate | Figure 9 AUC0-t bars | Sum | Results 3.5 AUC0-t | Figure 9 Cmax bars | Sum | Results 3.5 Cmax |
|---|---|---|---|---|---|---|
| LBW | 1.37, 0.82 | 2.19 | 2.19% | 13.29, 1.15 | 14.44 | 14.44% |
| DBIL | 12.73, 24.03 | 36.76 | 36.76% | 16.70, 14.84 | 31.54 | 31.54% |
| BMI | 14.88, 19.79 | 34.67 | 34.67% | 3.59, 3.69 | 7.28 | 7.28% |

The paper does not print the 5th and 95th percentile covariate *values*, so the
**observed minimum and maximum from Table 1 are used as proxies** (with n = 25
episodes, the 5th and 95th percentiles sit essentially at the extremes). AUC0-t
is computed on the paper's own measurement grid, truncated at the last sample
(97.5 h after dose start).

```{r figure-9}
grid_paper <- c(0, 0.75, 1.5, 1.5 + c(0.5, 1, 2, 4, 8, 12, 24, 48, 72, 96))

exposure_at <- function(BMI = med$BMI, DBIL = med$DBIL, LBM = med$LBM) {
  s <- rxode2::rxSolve(
    mod,
    rxode2::et(amt = dose_typ, dur = 1.5, cmt = "central") |> rxode2::et(grid_paper),
    params = c(BMI = BMI, DBIL = DBIL, LBM = LBM),
    omega = NA, returnType = "data.frame"
  )
  c(cmax = max(s$Cc), auct = trapz(s$time, s$Cc))
}

ref_exp <- exposure_at()

fig9 <- tibble::tribble(
  ~Covariate, ~Value, ~Percentile, ~`Printed AUC0-t`, ~`Printed Cmax`,
  "BMI",  18.64, "5th",  19.79,  3.69,
  "BMI",  28.97, "95th", -14.88, -3.59,
  "DBIL", 1.50,  "5th",  -12.73, -16.70,
  "DBIL", 7.90,  "95th", 24.03,  14.84,
  "LBM",  32.09, "5th",  1.37,   -13.29,
  "LBM",  59.40, "95th", 0.82,   -1.15
) |>
  rowwise() |>
  mutate(
    res = list(switch(
      Covariate,
      BMI  = exposure_at(BMI = Value),
      DBIL = exposure_at(DBIL = Value),
      LBM  = exposure_at(LBM = Value)
    )),
    `Model AUC0-t` = 100 * (res[["auct"]] / ref_exp[["auct"]] - 1),
    `Model Cmax`   = 100 * (res[["cmax"]] / ref_exp[["cmax"]] - 1)
  ) |>
  ungroup() |>
  select(Covariate, Value, Percentile,
         `Printed AUC0-t`, `Model AUC0-t`, `Printed Cmax`, `Model Cmax`)

fig9 |>
  knitr::kable(
    digits = 2,
    caption = paste(
      "Percent change in exposure versus the typical subject, model against the",
      "values printed inside the Figure 9 panels. Signs on the printed Cmax and",
      "AUC0-t bars follow the direction of change, which the tornado-plot axis",
      "does not itself carry."
    )
  )
```

The BMI and DBIL AUC0-t bars are reproduced to within about a percentage point,
which is a genuine test of the exponents 0.78 and -0.24 together with the
normalizing medians 23.44 and 2.82: a transcription error in any of the four
moves these numbers by far more than that.

```{r gate-figure-9}
auc_bars <- fig9 |> filter(Covariate %in% c("BMI", "DBIL"))

# Every quantity gated here is a TYPICAL-VALUE solve with `omega = NA`, so
# there is no cohort draw and no thread-count dependence -- these bounds are
# reproducible, and are set with roughly 2x headroom over the realised
# deviations recorded below rather than at the realised values themselves.
stopifnot(
  # Realised max 1.61 pp. A typo in either exponent (0.78, -0.24) or either
  # normalizing median (23.44, 2.82) moves these bars by 10 pp or more.
  max(abs(auc_bars$`Model AUC0-t` - auc_bars$`Printed AUC0-t`)) < 4,
  # LBM cannot change AUC0-inf at all, and changes AUC0-t only through the
  # shape of the truncated tail. Realised magnitudes 1.78 and 0.66 pp against
  # printed 1.37 and 0.82; the paper's own finding is "little (2.19%)".
  # Attaching any LBM exponent to CL instead would push this past 10 pp.
  max(abs(fig9$`Model AUC0-t`[fig9$Covariate == "LBM"])) < 3,
  # Realised max 2.97 pp on Cmax, which is a single point on a very steep
  # post-infusion decline (see the note below on grid sensitivity).
  max(abs(fig9$`Model Cmax` - fig9$`Printed Cmax`)) < 5,
  # Direction is gated only where the printed effect is materially non-zero.
  # The LBM 95th-percentile Cmax bar is a ~1% near-cancellation of two
  # opposing terms, so its sign is not diagnostic of anything; see below.
  all(with(
    fig9[abs(fig9$`Printed Cmax`) >= 3, ],
    sign(`Model Cmax`) == sign(`Printed Cmax`)
  ))
)
```

Note the sign pattern for LBM on Cmax, which is not the naive one. A *lower*
LBM shrinks `V1` (exponent +1.18), which alone would raise Cmax -- but it also
raises `CL2` steeply (exponent -2.55: at LBM 32.09 kg, `CL2` is about 2.9 times
its typical value) and raises `V2`, so drug is pulled into the shallow
peripheral compartment during the infusion and the peak falls instead. The
model reproduces that inversion, which is only possible if all three LBM
exponents are attached to the right parameters.

At the *high* end of LBM those two mechanisms very nearly cancel: `V1` grows by
24% (pushing Cmax down) while `CL2` falls to 63% of typical and `V2` to 77%
(pushing it up). The model nets +1.0% where Figure 9 prints 1.15% -- the same
magnitude with the opposite sign. This is the one bar in Figure 9 where the
model and the paper disagree in direction, and it is not diagnostic: the effect
is a ~1% residue of two effects an order of magnitude larger, so its sign is
set by whichever cancels marginally harder and is not stable against rounding
of the published exponents to two decimal places. The direction gate below is
therefore applied only to bars whose printed magnitude is at least 3%; the
magnitude of this bar is still gated.

# Typical-subject exposure versus Results 3.5

Results 3.5 reports that the simulated typical subject at 25 mg/m^2 had
AUC0-t = 9909.74 ng*h/mL and Cmax = 1824.91 ng/mL.

```{r typical-exposure}
tibble::tibble(
  Quantity = c("AUC0-t (ng*h/mL)", "Cmax (ng/mL)"),
  Paper = c(9909.74, 1824.91),
  Model = c(ref_exp[["auct"]], ref_exp[["cmax"]])
) |>
  mutate(`% diff` = 100 * (Model / Paper - 1)) |>
  knitr::kable(digits = 2, caption = "Typical subject at 25 mg/m^2 (median covariates).")
```

These are reproduced to about +6% on AUC0-t and +12% on Cmax, and the residual
is a **simulation-design** gap rather than a model-parameter one. Two unstated
inputs drive it, and they are documented as deviations rather than tuned away:

* **The dose in mg is not stated.** 25 mg/m^2 must be multiplied by a body
  surface area, and the paper does not say which one it used. The median BSA
  from Table 1 (1.678 m^2) is used here, giving 41.95 mg. Because the model is
  linear, AUC0-t scales exactly with dose, so a BSA of about 1.59 m^2 would
  reproduce the AUC0-t figure exactly. No BSA reproduces both numbers, so the
  gap is not purely a dose choice.
* **Cmax is read off a grid.** The decline immediately after end of infusion is
  extremely steep (concentration falls by about 60% in the following 30 min,
  since `CL3 / V1` is roughly 2.2 /h). Sampling the peak even 3 min after
  1.5 h drops it to about 1855 ng/mL, essentially the published value. A
  simulation output grid that does not land exactly on the end of infusion
  therefore understates Cmax by roughly the observed amount.

Consistent with that reading, the *relative* covariate effects in Figure 9 --
which are insensitive to both the absolute dose and the peak grid -- are
reproduced to about a percentage point above.

# Virtual cohort

Original subject-level data are not public. The cohort below matches the
Table 1 marginal distributions: each covariate is drawn log-normal with the
published median and interquartile range, then truncated to the published
range. The paper's collinearity screen retained BMI, DBIL and LBM precisely
because they were mutually uncorrelated (r < 0.5, Results 3.3), so drawing them
independently is consistent with the source; BSA is drawn independently too and
enters only through the mg/m^2 to mg dose conversion.

```{r cohort}
# set.seed() seeds R's RNG for the covariate draws below. It does NOT seed
# rxode2's simulation RNG, whose streams are partitioned per solver thread, so
# the etas differ between a 16-thread workstation and a 2-core CI runner. Every
# assertion on a cohort quantity below is written to hold for any such draw.
set.seed(20230116)

n_per_arm <- 100L
dose_levels <- c(2, 4, 6, 10, 14, 30, 45)  # mg/m^2, Table 2 cycle 1

# Log-normal from a published median and IQR: sigma = log(q75/q25) / (2*0.6745).
rlnorm_iqr <- function(n, median, q25, q75, lo, hi) {
  sigma <- log(q75 / q25) / (2 * stats::qnorm(0.75))
  pmin(pmax(stats::rlnorm(n, log(median), sigma), lo), hi)
}

make_arm <- function(dose_mgm2, n, id_offset) {
  subj <- tibble::tibble(
    id   = id_offset + seq_len(n),
    BMI  = rlnorm_iqr(n, 23.44, 21.05, 24.91, 18.64, 28.97),
    DBIL = rlnorm_iqr(n, 2.82, 2.30, 4.00, 1.50, 7.90),
    LBM  = rlnorm_iqr(n, 49.53, 38.36, 55.45, 32.09, 59.40),
    BSA  = rlnorm_iqr(n, 1.678, 1.600, 1.841, 1.420, 1.938),
    treatment = sprintf("%g mg/m^2", dose_mgm2)
  ) |>
    mutate(amt_mg = dose_mgm2 * BSA)

  dosing <- subj |>
    mutate(time = 0, amt = amt_mg, evid = 1L, dur = 1.5, cmt = "central")
  obs <- subj |>
    tidyr::crossing(time = grid_paper) |>
    mutate(amt = NA_real_, evid = 0L, dur = NA_real_, cmt = "central")

  bind_rows(dosing, obs) |>
    select(id, time, amt, evid, dur, cmt, treatment, BMI, DBIL, LBM, BSA, amt_mg) |>
    arrange(id, time, desc(evid))
}

events <- do.call(bind_rows, lapply(seq_along(dose_levels), function(i) {
  make_arm(dose_levels[i], n_per_arm, id_offset = (i - 1L) * n_per_arm)
}))

# Disjoint IDs across arms -- duplicate IDs would silently merge subjects.
stopifnot(
  !anyDuplicated(unique(events[, c("id", "time", "evid")])),
  length(unique(events$id)) == n_per_arm * length(dose_levels)
)
```

```{r simulate}
# rxSolve on an rxUi scales super-linearly in subjects per call, so solve one
# arm at a time rather than passing all 700 subjects in a single call.
sim <- do.call(bind_rows, lapply(split(events, events$treatment), function(ev_arm) {
  rxode2::rxSolve(
    mod, ev_arm,
    keep = c("treatment", "BMI", "DBIL", "LBM", "amt_mg"),
    returnType = "data.frame"
  )
})) |>
  mutate(treatment = factor(treatment, levels = sprintf("%g mg/m^2", dose_levels)))

stopifnot(nrow(sim) > 0, all(sim$Cc >= 0), !anyNA(sim$Cc))
```

## Replicating Figure 1 (mean concentration-time profiles by dose level)

```{r figure-1}
sim |>
  filter(time > 0) |>
  group_by(treatment, time) |>
  summarise(Cc = mean(Cc), .groups = "drop") |>
  ggplot(aes(time, Cc, colour = treatment)) +
  geom_line() +
  geom_point(size = 1) +
  scale_y_log10() +
  labs(
    x = "Time since start of infusion (h)", y = "TQ-B3203 (ng/mL)",
    colour = "Dose level",
    title = "Mean TQ-B3203 plasma concentration by dose level",
    caption = "Replicates Figure 1 of Li 2023 (semilogarithmic mean profiles, cycle 1)."
  )
```

The profiles are parallel across dose levels -- the model is linear, and Results
3.2 reports no accumulation and a biphasic decline (a rapid then a slow
elimination phase), which the three-compartment structure produces.

## Replicating Figure 7 (visual predictive check)

```{r figure-7}
sim |>
  filter(time > 0) |>
  group_by(treatment, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05), Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~treatment, scales = "free_y") +
  scale_y_log10() +
  labs(
    x = "Time since start of infusion (h)", y = "TQ-B3203 (ng/mL)",
    title = "Simulated 5th / 50th / 95th percentiles by dose level",
    caption = "Structure of Figure 7 of Li 2023 (VPC of log-transformed observations)."
  )
```

## Cohort-level closed-form check

The `Dose / CL` identity holds per subject, so it can be re-run across the whole
cohort: each subject's own drawn `cl` is compared against their own solved
profile. Here the profile is integrated on the paper's own sparse 13-point
grid rather than a dense one, so the comparison also quantifies how much
approximation the published sampling schedule carries.

```{r gate-cohort-aucinf}
per_subject <- sim |>
  group_by(id, treatment) |>
  summarise(
    cl = dplyr::first(cl), amt_mg = dplyr::first(amt_mg),
    auct = trapz(time, Cc), cmax = max(Cc), tmax = time[which.max(Cc)],
    .groups = "drop"
  ) |>
  mutate(
    auc_closed = 1000 * amt_mg / cl,
    auct_frac  = auct / auc_closed
  )

stopifnot(
  # Structural, not statistical: the peak of a linear model under zero-order
  # input is at end of infusion for EVERY subject whatever their covariates,
  # so all() is the right quantifier here.
  all(abs(per_subject$tmax - 1.5) < 1e-8),
  # This ratio is a cohort quantity, so it is gated on its centre and on a
  # robust quantile rather than on either extreme (which is one draw on one
  # thread count). Realised median 1.012, 90th percentile of |ratio - 1|
  # about 0.04, full range 0.885-1.048 over 700 subjects.
  abs(median(per_subject$auct_frac) - 1) < 0.10,
  quantile(abs(per_subject$auct_frac - 1), 0.90) < 0.15
)

per_subject |>
  group_by(treatment) |>
  summarise(
    `Median` = median(auct_frac), `5th pct` = quantile(auct_frac, 0.05),
    `95th pct` = quantile(auct_frac, 0.95), .groups = "drop"
  ) |>
  knitr::kable(
    digits = 4,
    caption = paste(
      "Linear-trapezoid AUC0-t on the paper's 13-point grid, divided by the",
      "closed-form Dose/CL. Two errors of opposite sign are in play: the grid",
      "truncates at 97.5 h (biasing the ratio down) while a linear trapezoid",
      "over a convex decline overestimates each segment (biasing it up), so",
      "the ratio sits slightly above 1 rather than below it."
    )
  )
```

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time = 0 row per subject; TQ-B3203 is given intravenously to
# treatment-naive patients, so the pre-dose concentration is 0.
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(sim_nca), Cc ~ time | treatment + id)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE,
  half.life = TRUE, cl.obs = TRUE, vz.obs = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_df <- as.data.frame(nca_res)

# UNIT RESCALE, mandatory before any comparison of CL or Vz against the paper.
# PKNCA divides dose by exposure in whatever units it was handed. Doses here
# are mg and concentrations ng/mL, so cl.obs comes back as mg/(ng*h/mL) and
# vz.obs as mg/(ng/mL). Since 1 mg = 1e6 ng, both are the L/h and L value
# divided by 1000. Li 2023 Table 2 reports CL in L/h and Vz in L, so rescale
# here rather than at each comparison site.
nca_df <- nca_df |>
  mutate(PPORRES = ifelse(PPTESTCD %in% c("cl.obs", "vz.obs"),
                          PPORRES * 1000, PPORRES))

stopifnot(
  nrow(nca_df) > 0,
  # Guard the rescale itself: a plausible CL is order 1-10 L/h, not 1e-3.
  median(nca_df$PPORRES[nca_df$PPTESTCD == "cl.obs"]) > 1
)
```

```{r pknca-sanity}
# PKNCA's CL must recover the model's own CL, and its AUCinf the closed form.
nca_cl <- nca_df |>
  filter(PPTESTCD == "cl.obs") |>
  select(id, treatment, cl_nca = PPORRES) |>
  left_join(per_subject |> select(id, treatment, cl), by = c("id", "treatment")) |>
  mutate(pct = 100 * (cl_nca / cl - 1))   # cl_nca already rescaled to L/h above

stopifnot(
  nrow(nca_cl) == nrow(per_subject),
  !anyNA(nca_cl$pct),
  # Both sides use the same drawn parameters, so the residual is purely
  # log-linear terminal-slope extrapolation error on a sparsely sampled
  # 3-compartment profile. Realised median +1.23%, 90th percentile of the
  # absolute difference 2.50%, worst subject 7.74% over 700 subjects; these
  # are cohort statistics so the centre and a robust quantile are gated, not
  # the maximum.
  abs(median(nca_cl$pct)) < 4,
  quantile(abs(nca_cl$pct), 0.90) < 6
)

cat(sprintf(
  "PKNCA CL vs model CL: median %+.2f%%, 90th pct of |diff| %.2f%%\n",
  median(nca_cl$pct), quantile(abs(nca_cl$pct), 0.90)
))
```

## Comparison against the published NCA (Table 2)

Table 2 reports observed NCA by dose level for cycle 1. The comparison below is
inherently coarse: each published cell is the mean of **1 to 3 patients** with
their own covariate values, whereas the simulated column is the median of 100
virtual subjects drawn from the Table 1 marginals. Dose-level means built on
n = 1 carry no usable precision, so the table is read for dose-proportionality
and for agreement on `CL` and half-life rather than cell by cell.

```{r nca-compare}
published <- tibble::tribble(
  ~treatment,      ~cmax, ~tmax, ~auclast, ~aucinf.obs, ~half.life, ~cl.obs, ~vz.obs,
  "2 mg/m^2",         85,   1.5,      775,         801,       10.3,     4.6,    68.1,
  "4 mg/m^2",        284,   1.5,     1410,        1423,       11.3,     4.7,    77.3,
  "6 mg/m^2",        533,   1.5,     2705,        2778,       16.4,     4.0,    90.0,
  "10 mg/m^2",      1044,   1.5,     5258,        5326,       18.0,     3.4,    87.4,
  "14 mg/m^2",      1559,   1.5,     8584,        8731,       21.6,     2.9,    90.0,
  "30 mg/m^2",      2357,   1.5,    15882,       16602,       33.3,     2.8,   133.8,
  "45 mg/m^2",      4956,   1.5,    26251,       27026,       34.3,     2.9,   148.7
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_df,
  reference = as.data.frame(published),
  by = "treatment",
  units = c(cmax = "ng/mL", auclast = "ng*h/mL", aucinf.obs = "ng*h/mL",
            tmax = "h", half.life = "h", cl.obs = "L/h", vz.obs = "L"),
  tolerance_pct = 20
)

knitr::kable(
  cmp, digits = 3,
  caption = paste(
    "Simulated (median of 100 virtual subjects per arm) versus Li 2023 Table 2",
    "observed means (n = 1 to 3 patients per dose level).",
    "* differs from the reference by more than 20%.",
    "The clearance and volume rows are labelled CL/F and Vz/F by the shared",
    "label helper; TQ-B3203 is given intravenously, so they are simply CL",
    "and Vz."
  )
)
attr(cmp, "footnote")
```

```{r nca-compare-gates}
# Dose proportionality is a property of the linear model and is checked
# directly rather than through the small-n published cells.
dp <- per_subject |>
  mutate(dose_mgm2 = as.numeric(sub(" .*", "", as.character(treatment)))) |>
  group_by(dose_mgm2) |>
  summarise(auc_per_mg = median(auct / amt_mg), .groups = "drop")

stopifnot(
  # A linear model gives dose-normalised AUC independent of dose level. The
  # spread here is cohort sampling only, so bound the relative spread rather
  # than any pair of adjacent levels.
  diff(range(dp$auc_per_mg)) / median(dp$auc_per_mg) < 0.20
)

# CL: the paper's own NCA spans 2.8-6.0 L/h across dose levels and cycles
# (Table 2), against a model typical value of 3.97 L/h. Assert the cohort
# median sits inside the published envelope rather than matching any one cell.
cl_med <- median(per_subject$cl)
stopifnot(cl_med > 2.8, cl_med < 6.0)

# Terminal half-life: Results 3.2 reports 10-41 h (median 19.62, mean 20).
hl <- nca_df |> filter(PPTESTCD == "half.life") |> pull(PPORRES)
stopifnot(!anyNA(hl), median(hl) > 8, median(hl) < 41)

cat(sprintf(
  "Dose-normalised AUC0-t spread %.1f%% of median | cohort median CL %.2f L/h (paper NCA 2.8-6.0) | median t1/2 %.1f h (paper 10-41, median 19.62)\n",
  100 * diff(range(dp$auc_per_mg)) / median(dp$auc_per_mg), cl_med, median(hl)
))
```

Many rows are starred, and they are starred for one coherent reason that is
worth setting out, because it is a property of the **published model** rather
than of this extraction.

## The published NCA is dose-dependent; the published model is not

Table 2 can be interrogated on its own, with no model at all. Doing so shows
that the observed non-compartmental summary is not dose-linear:

```{r dose-linearity-published}
published_trend <- published |>
  mutate(dose_mgm2 = as.numeric(sub(" .*", "", treatment))) |>
  transmute(
    `Dose (mg/m^2)` = dose_mgm2,
    `AUC0-t / dose` = auclast / dose_mgm2,
    `Cmax / dose`   = cmax / dose_mgm2,
    `CL (L/h)`      = cl.obs,
    `t1/2 (h)`      = half.life
  )

rho_hl <- cor(published_trend$`Dose (mg/m^2)`, published_trend$`t1/2 (h)`,
              method = "spearman")
rho_cl <- cor(published_trend$`Dose (mg/m^2)`, published_trend$`CL (L/h)`,
              method = "spearman")

# Facts about the printed table: no simulation, no sampling noise. Tested with
# a tolerance rather than `== 1` because a rank correlation of one is not
# guaranteed to come back bit-exact.
stopifnot(
  rho_hl > 0.999,
  rho_cl < -0.7,
  max(published_trend$`CL (L/h)`) / min(published_trend$`CL (L/h)`) > 1.5
)

knitr::kable(
  published_trend, digits = 1,
  caption = sprintf(
    paste("Li 2023 Table 2 cycle-1 values, dose-normalised. Spearman rank",
          "correlation with dose: t1/2 %+.3f, CL %+.3f."),
    rho_hl, rho_cl
  )
)
```

Observed terminal half-life increases **monotonically across all seven dose
levels** (Spearman `r sprintf("%+.3f", rho_hl)`), observed clearance falls by a
factor of `r sprintf("%.2f", max(published_trend[["CL (L/h)"]]) / min(published_trend[["CL (L/h)"]]))`,
and dose-normalised AUC0-t spans a 1.7-fold range. The final model, however, has
first-order elimination and no dose term, so it predicts a dose-**independent**
half-life and clearance. It therefore cannot track those trends, and the starred
`t1/2`, `CL`, `Vz`, `AUC` and high-dose `Cmax` rows above all follow from that
single mismatch.

The most obvious benign explanation is that the assay, linear over 0.5-500
ng/mL (Methods 2.3), truncates the terminal phase early at low doses and biases
the fitted terminal slope steep. That explanation is testable, and it does not
survive the test:

```{r lloq-test}
nca_medians <- function(lloq) {
  sn <- sim |>
    filter(!is.na(Cc)) |>
    mutate(Cc = ifelse(time > 0 & Cc < lloq, NA_real_, Cc)) |>
    filter(!is.na(Cc)) |>
    select(id, time, Cc, treatment)
  sn <- bind_rows(
    sn, sn |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
  ) |>
    distinct(id, treatment, time, .keep_all = TRUE) |>
    arrange(id, treatment, time)

  d <- PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(as.data.frame(sn), Cc ~ time | treatment + id),
    dose_obj,
    intervals = data.frame(start = 0, end = Inf, half.life = TRUE, cl.obs = TRUE)
  )
  as.data.frame(PKNCA::pk.nca(d)) |>
    mutate(PPORRES = ifelse(PPTESTCD == "cl.obs", PPORRES * 1000, PPORRES)) |>
    group_by(treatment, PPTESTCD) |>
    summarise(m = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
    tidyr::pivot_wider(names_from = PPTESTCD, values_from = m) |>
    mutate(dose_mgm2 = as.numeric(sub(" .*", "", as.character(treatment)))) |>
    arrange(dose_mgm2)
}

lloq_free <- nca_medians(0)
lloq_real <- nca_medians(0.5)

hl_span <- function(x) diff(range(x$half.life))
comparison <- tibble::tibble(
  Scenario = c("Simulated, no LLOQ", "Simulated, LLOQ 0.5 ng/mL",
               "Li 2023 Table 2 (observed)"),
  `t1/2 min (h)` = c(min(lloq_free$half.life), min(lloq_real$half.life),
                     min(published$half.life)),
  `t1/2 max (h)` = c(max(lloq_free$half.life), max(lloq_real$half.life),
                     max(published$half.life)),
  `t1/2 span (h)` = c(hl_span(lloq_free), hl_span(lloq_real),
                      diff(range(published$half.life))),
  `CL min (L/h)` = c(min(lloq_free$cl.obs), min(lloq_real$cl.obs),
                     min(published$cl.obs)),
  `CL max (L/h)` = c(max(lloq_free$cl.obs), max(lloq_real$cl.obs),
                     max(published$cl.obs))
)

knitr::kable(
  comparison, digits = 2,
  caption = paste(
    "Applying the paper's own assay lower limit to the simulated profiles does",
    "not reproduce the observed dose trend in half-life or clearance."
  )
)
```

```{r lloq-test-gates}
stopifnot(
  # The model is flat in dose whether or not the assay limit is applied: the
  # per-arm median half-life spans only a few hours against the observed 24 h.
  # These are per-arm medians of 100 subjects, so bounded with headroom over
  # the realised spans (2.4 h without LLOQ, 3.1 h with).
  hl_span(lloq_free) < 8,
  hl_span(lloq_real) < 8,
  # Censoring at the real LLOQ does not close even a third of the gap.
  hl_span(lloq_real) < 0.35 * diff(range(published$half.life)),
  # Simulated CL stays within ~10% of the model typical value at every dose
  # level, against the 1.68-fold observed span.
  max(lloq_real$cl.obs) / min(lloq_real$cl.obs) < 1.25
)
```

Censoring at the real assay limit moves the simulated half-life span from about
2 h to about 3 h, against an observed span of 24 h, and leaves simulated
clearance essentially unchanged. So the dose trend in Table 2 is not an assay
artifact, and the linear model does not reproduce it.

Two readings are available and the paper does not settle between them. The
trend may be real saturation of elimination, in which case the published model
is misspecified over the 2-45 mg/m^2 range. Or it may be confounding: this is a
dose-**escalation** design, so each dose level is a different set of one to
three patients, and clearance in this model depends on BMI and direct
bilirubin, so a drift in cohort composition across the escalation would produce
exactly this pattern without any nonlinearity. The authors' own stepwise
covariate search never tested dose as a covariate, and with n = 15 patients the
per-level cells (n = 1 at three of the seven levels) carry no usable precision
either way.

Nothing here is tuned. The model reproduces every *model-derived* published
quantity -- the closed-form clearance identities exactly, and the Figure 9
covariate sensitivities to about a percentage point -- and the disagreement is
confined to observed quantities the published linear model was never able to
describe. That is recorded rather than smoothed over.

# Assumptions and deviations

* **IIV on V1 is declared but held at zero.** Equation 9 prints
  `V1 = 4.81 * (LBW/49.53)^1.18 * exp(eta_V1)`, but Table 4 reports omega^2 for
  only CL, V3, CL2 and CL3 -- there is no omega^2 for V1 anywhere in the paper.
  Rather than invent a variance, `etalvc` is declared and wrapped in `fixed(0)`
  so the printed structure stays visible and a downstream re-fit can free it.
  V2 gets no eta at all, because no equation or table row in the paper shows
  one for it.
* **No IIV correlation structure.** Methods 2.5.2 says the correlation between
  IIVs "should be clarified to construct a covariance model", but no covariance
  or correlation is tabulated, so the four random effects are independent
  diagonal terms.
* **The residual error is encoded as log-normal, not proportional.** Results
  3.3 selected "a log-additive model" for intra-individual variability and
  estimated on log-transformed concentrations. A Phoenix NLME log-additive
  residual is `log(C_obs) = log(C_pred) + eps`, which is exactly
  `Cc ~ lnorm(expSd)` with `expSd = 0.200`. Encoding it as `prop(propSd)`
  would be a different error structure.
* **The dose for the Figure 9 typical subject is derived, not published.**
  25 mg/m^2 needs a BSA to become an absolute mg amount and the paper does not
  state which one it used; the Table 1 median BSA of 1.678 m^2 is used, giving
  41.95 mg. See the typical-exposure section for why no single BSA reproduces
  both published numbers.
* **Figure 9's 5th and 95th percentile covariate values are not published**, so
  the Table 1 observed minimum and maximum are used as proxies. With n = 25
  episodes these are close to the true percentiles, and the resulting AUC0-t
  changes land within about a percentage point of the printed bars, which
  supports the substitution.
* **Known deviation, not gated: absolute typical Cmax and AUC0-t.** The model
  gives about +12% on Cmax and +6% on AUC0-t against Results 3.5. Both are
  attributable to unstated simulation-design inputs (the BSA behind the mg dose
  and the output grid resolution at the peak) and neither is reproducible by any
  change that keeps the printed parameters, so no parameter was adjusted. The
  numbers are rendered above so the gap stays visible.
* **The LBM body-composition formula is not stated.** Li 2023 cites Park 2018
  as the reason lean body weight was screened but never names the formula
  (James, Boer and Hume differ by several kg at a given height and weight,
  which rescales a median-normalized power effect). A downstream user should
  check that their formula reproduces the Table 1 distribution: median 49.53
  kg, range 32.09-59.40 kg, for a cohort of median weight 64.0 kg and height
  164 cm.
* **Covariates are drawn independently.** Table 1 gives only marginal medians,
  IQRs and ranges, and Figure 3 shows a correlation matrix without printed
  coefficients. The paper's collinearity screen retained BMI, DBIL and LBM
  because they were mutually uncorrelated (r < 0.5), so independent draws are
  consistent with the source. BSA is also drawn independently and affects only
  the dose conversion.
* **Covariates are time-fixed at baseline**, matching a single-dose analysis
  with baseline laboratory values.
* **No inter-occasion variability.** IOV on CL and V1 was tested and rejected
  (Results 3.3): the log-additive error moved only from 22.4% to 21.4% and fit
  diagnostics stayed poor, and no cycle 1 versus cycle 2 difference was found.
* **`CRCL` in Table 1 carries a unit typo in the source.** The paper labels
  creatinine clearance "mg/dL" for both CLcr and adjusted CLcr; creatinine
  clearance is a flow, conventionally mL/min, and the Cockcroft-Gault formula
  in Eq. 4 returns mL/min. The covariate was screened and not retained, so this
  does not affect the model.
* **Erratum search: negative.** EuropePMC returns an empty
  `commentCorrectionList` for DOI 10.3389/fphar.2023.1102244, and the PubMed
  record (PMID 36726585) carries no `CommentsCorrections` element and a sole
  publication type of "Journal Article" -- so no erratum, corrigendum,
  expression of concern or retraction is registered against this article by
  either index.
* **Supplement: retrieved and confirmed parameter-free.** The single
  supplementary file (`DataSheet1.DOCX`, via the EuropePMC
  `supplementaryFiles?includeInlineImage=false` route) contains exactly three
  figures and no tables: S1, liposome size stability over 3 months at 4 C and
  over 7 days in PBS, plasma and fetal bovine serum; S2, whole-body
  near-infrared biodistribution imaging of probe-embedded TLI in HT-29
  tumour-bearing mice; S3, the chemical structures of TQ-B3203 and its
  deuterated internal standard TQ-B3203-d8. The paper cites only these three
  figures and no supplementary table anywhere in its text, so the complete
  final model is contained in Table 4 with Equations 7, 9 and 10 of the main
  article, and nothing in this extraction depends on the supplement.
