---
title: "Tedizolid (Li 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tedizolid (Li 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Li_2021_tedizolid"))$reference`
* Description: `r rxode2::rxode(readModelDb("Li_2021_tedizolid"))$description`
* Article: <https://doi.org/10.1128/AAC.00895-21> (open access; the supplement
  holds the per-trial demographics, Tables A1-A3)
* Backbone model: Flanagan S et al. (2014), <https://doi.org/10.1128/AAC.03423-14>

Tedizolid phosphate is an oxazolidinone prodrug that nonspecific phosphatases
hydrolyse to tedizolid, the active moiety. The model describes plasma tedizolid.

# Population

Li 2021 updated the earlier tedizolid population PK model (Flanagan 2014) with
the phase 3 trial PN012 in adolescents (12 to < 18 years) with acute bacterial
skin and skin-structure infection (ABSSSI) and the phase 1 trial PN013 in
hospitalized children aged 2 to < 12 years. The final data set pooled 16 trials,
1,312 participants and 9,756 plasma concentrations: 945 adults with ABSSSI,
223 healthy participants, 41 hospitalized children and adolescents with
suspected Gram-positive infection, and 103 adolescents with ABSSSI (91 from
PN012). Age ranged from 3 to 94 years (median 40) and body weight from 12.6 to
226 kg (median 76.0 kg); 32.6% were female; 70.9% were White, 15.2% Asian,
12.1% Black and 1.8% other; 7.7% had diabetes (Results, "Participants";
supplement Tables A1 and A2). 5,146 samples followed oral and 4,647 followed
intravenous dosing (supplement Table A3).

The same information is available programmatically:

```{r population}
str(readModelDb("Li_2021_tedizolid")()$population)
```

# Source trace

Every `ini()` value carries an in-file comment pointing at its source. They are
collected here.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Two-compartment disposition, linear elimination | -- | Results "PopPK analysis"; Methods "PopPK model development" |
| Sigmoidal oral absorption: zero-order release into depot, then first-order `ka` | -- | Methods "PopPK model development"; Flanagan 2014 Results |
| `CL = TVCL * (WT/77.3)^thetaWT * (1 + thetaInfec * INFEC) * exp(eta)` | -- | Table 1 footnote b |
| `Vc = TVVc * (WT/77.3)^thetaWT * (1 + thetaInfec * INFEC) * (1 + thetaDiab * DIAB) * exp(eta)` | -- | Table 1 footnote b |
| `Vp = TVVp * (WT/77.3)^thetaWT * exp(eta)` | -- | Table 1 footnote b |
| `ld2` (intravenous infusion time) | log(0.810 h), fixed | Table 1 "Infusion time, fixed" |
| `lfdepot` | log(0.857) | Table 1 "F1" |
| `ld1` (oral zero-order duration) | log(0.175 h) | Table 1 "Zero-order duration" |
| `lka` | log(1.47 1/h) | Table 1 "Ka" |
| `ltlag` | log(0.226 h) | Table 1 "Lag time" |
| `lcl` | log(5.39 L/h) | Table 1 "CL" |
| `e_wt_cl` (CL and Q) | 0.408 | Table 1 CL "wt (power model)"; Q "Same as for CL" |
| `e_csssi_cl` | 0.220 | Table 1 CL "Infection (%)" = 22.0 |
| `lvc` | log(58.5 L) | Table 1 "Vc" |
| `e_wt_vc` | 0.903 | Table 1 Vc "wt (power model)" |
| `e_csssi_vc` | 0.0987 | Table 1 Vc "Infection (%)" = 9.87 |
| `e_dis_diab_vc` | -0.143 | Table 1 Vc "Diabetes (%)" = -14.3 |
| `lq` | log(1.43 L/h) | Table 1 "Q" |
| `lvp` | log(15.6 L) | Table 1 "Vp" |
| `e_wt_vp` | 0.678 | Table 1 Vp "wt (power model)" |
| `etald2` | 0.00680, fixed (8.26% CV) | Table 1 "Infusion time" IIV |
| `etald1` | 2.036 (258% CV) | Table 1 "Zero-order duration" IIV |
| `etalka` | 0.4656 (77.0% CV) | Table 1 "Ka" IIV |
| `etaltlag` | 0.6931 (100% CV) | Table 1 "Lag time" IIV |
| `etalcl`, `etalvc` block | 0.09865, 0.04840, 0.06157 | Table 1 CL 32.2% CV, Vc 25.2% CV, correlation 62.1% |
| `etalvp` | 0.02466 (15.8% CV) | Table 1 "Vp" IIV |
| `expSd` | 0.123 | Table 1 "RV for non-phase 3 trials (%)" |
| `e_study_phase3_expsd` | 4.92 | Table 1 "RV for study 104 and phase 3 trials (fold)", footnote h |
| `e_route_oral_expsd` | 2.01 | Table 1 "RV for oral data (fold)", footnote h |

# Dose basis

The paper gives doses as 200 mg of tedizolid phosphate but does not say whether
the analysis data set carried the dose as prodrug mass or as tedizolid. The
model is dosed in tedizolid free-base equivalents, the molar equivalent of the
prodrug dose. The weight-quartile exposures below reproduce Table 3 on that
basis and are about 22% too high if the prodrug mass is used instead.

```{r dose-basis}
mw_tedizolid <- 370.34 # C17H15FN6O3
mw_tedizolid_phosphate <- 450.32 # C17H16FN6O6P
dose_phosphate_mg <- 200
dose_mg <- dose_phosphate_mg * mw_tedizolid / mw_tedizolid_phosphate
dose_mg
```

# Structural checks on the typical individual

Single doses to a typical 77.3 kg healthy participant, with all random effects
set to zero. For a linear model, `CL * AUC(0-inf)` must return the dose that
reached the circulation: the full dose intravenously and `F * dose` orally.
Intravenous records carry `rate = -2`. Without it rxode2 ignores
`dur(central)`, gives a bolus, and moves only Cmax, which no AUC check can see.
So Tmax is also required to land at the end of the modelled 0.81 h infusion.

```{r typical-checks}
mod <- readModelDb("Li_2021_tedizolid")
mod_typ <- rxode2::zeroRe(rxode2::rxode(mod))

typ_cov <- data.frame(
  WT = 77.3,
  DIS_CSSSI = 0L,
  DIS_DIAB = 0L,
  STUDY_PHASE3 = 0L,
  ROUTE_ORAL = 0L
)

single_dose_events <- function(route_cmt, id) {
  obs_times <- sort(unique(c(
    0,
    exp(seq(log(0.01), log(4), length.out = 120)),
    seq(4, 240, by = 0.5)
  )))
  dplyr::bind_rows(
    data.frame(
      id = id, time = 0, amt = dose_mg, evid = 1L,
      cmt = route_cmt, rate = -2
    ),
    data.frame(
      id = id, time = obs_times, amt = 0, evid = 0L,
      cmt = "central", rate = 0
    )
  ) |>
    dplyr::mutate(admin = ifelse(route_cmt == "depot", "Oral", "IV")) |>
    dplyr::arrange(time, dplyr::desc(evid))
}

ev_typ <- dplyr::bind_rows(
  single_dose_events("central", 1L),
  single_dose_events("depot", 2L)
)
ev_typ <- cbind(ev_typ, typ_cov[rep(1, nrow(ev_typ)), ])
ev_typ$ROUTE_ORAL <- as.integer(ev_typ$admin == "Oral")

sim_typ <- as.data.frame(rxode2::rxSolve(
  mod_typ, ev_typ,
  keep = "admin", returnType = "data.frame",
  atol = 1e-10, rtol = 1e-10
))

# All three ODE states must be integrated; an analytic shortcut that dropped
# the peripheral compartment would still pass the AUC identity.
stopifnot("peripheral1" %in% names(sim_typ))

conc_typ <- sim_typ |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, admin, time, Cc)
dose_typ <- ev_typ |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, admin, time, amt)

nca_typ <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_typ, Cc ~ time | admin + id),
  PKNCA::PKNCAdose(dose_typ, amt ~ time | admin + id),
  intervals = data.frame(
    start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
    aucinf.obs = TRUE, half.life = TRUE
  )
))

typ_res <- as.data.frame(nca_typ$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  dplyr::select(admin, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

cl_typ <- 5.39
f_typ <- 0.857
typ_res <- typ_res |>
  dplyr::mutate(
    expected_auc = ifelse(admin == "Oral", f_typ, 1) * dose_mg / cl_typ,
    auc_pct_diff = 100 * (aucinf.obs - expected_auc) / expected_auc
  )

typ_res |>
  dplyr::rename(
    "Route" = admin,
    "Cmax (ug/mL)" = cmax,
    "Tmax (h)" = tmax,
    "AUC0-inf (ug*h/mL)" = aucinf.obs,
    "t1/2 (h)" = half.life,
    "F * Dose / CL (ug*h/mL)" = expected_auc,
    "AUC % diff" = auc_pct_diff
  ) |>
  knitr::kable(digits = 3, caption = "Typical-value single-dose NCA (77.3 kg healthy participant).")

iv_row <- typ_res[typ_res$admin == "IV", ]
bolus_cmax <- dose_mg / 58.5
stopifnot(
  all(abs(typ_res$auc_pct_diff) < 1),
  # Peak at the end of the 0.81 h modelled infusion, not at time zero.
  abs(iv_row$tmax - 0.81) < 0.05,
  # A bolus would peak at Dose / Vc at time zero. About 5% of the dose is
  # eliminated or distributed during the 0.81 h infusion, so the infusion
  # peak sits a few percent below the bolus value.
  iv_row$cmax < 0.98 * bolus_cmax
)
```

# Adolescent exposure by weight quartile (Table 3)

Li 2021 Table 3 gives the geometric-mean AUC(0-24 h) and Cmax on day 1 and on
the last dosing day for the 91 PN012 adolescents, split into body-weight
quartiles. Those values are post hoc (empirical Bayes) estimates. They are
compared here with a virtual PN012 cohort: 200 adolescents with ABSSSI and no
diabetes, dosed with 200 mg tedizolid phosphate intravenously once daily for
6 days. Weights follow a log-normal distribution matched to the PN012 median of
56.6 kg, redrawn inside the observed 27.6-126 kg range (supplement Table A2).
Most PN012 samples followed intravenous dosing (350 of 430 evaluable samples,
supplement Table A3), so the cohort is dosed intravenously.

```{r cohort}
# rxSetSeed fixes rxode2's random stream per solver thread, not across thread
# counts or rxode2 versions, so every assertion below is on a centre or a
# group geometric mean, never on a single extreme subject.
rxode2::rxSetSeed(20211117)
set.seed(20211117)

draw_weights <- function(n, median_wt, sdlog, lower, upper) {
  out <- numeric(0)
  while (length(out) < n) {
    w <- exp(stats::rnorm(n, log(median_wt), sdlog))
    out <- c(out, w[w >= lower & w <= upper])
  }
  out[seq_len(n)]
}

n_per_arm <- 200
obs_grid <- sort(unique(c(
  seq(0, 24, by = 0.1),
  seq(120, 144, by = 0.1)
)))

multi_dose_events <- function(ids, weights, route_cmt, csssi, diab) {
  dose_rows <- expand.grid(time = 24 * (0:5), id = ids)
  obs_rows <- expand.grid(time = obs_grid, id = ids)
  ev <- dplyr::bind_rows(
    data.frame(
      id = dose_rows$id, time = dose_rows$time, amt = dose_mg,
      evid = 1L, cmt = route_cmt, rate = -2
    ),
    data.frame(
      id = obs_rows$id, time = obs_rows$time, amt = 0,
      evid = 0L, cmt = "central", rate = 0
    )
  )
  ev$WT <- weights[match(ev$id, ids)]
  ev$DIS_CSSSI <- csssi
  ev$DIS_DIAB <- diab[match(ev$id, ids)]
  ev$STUDY_PHASE3 <- 1L
  ev$ROUTE_ORAL <- as.integer(route_cmt == "depot")
  dplyr::arrange(ev, id, time, dplyr::desc(evid))
}

ado_ids <- seq_len(n_per_arm)
ado_wt <- draw_weights(n_per_arm, 56.6, 0.28, 27.6, 126)
ev_ado_iv <- multi_dose_events(ado_ids, ado_wt, "central", 1L, rep(0L, n_per_arm))
sim_ado_iv <- as.data.frame(rxode2::rxSolve(
  mod, ev_ado_iv,
  keep = "WT", returnType = "data.frame"
))

summary(ado_wt)
```

```{r nca-helper}
nca_day1_last <- function(sim, ev, group_col) {
  conc <- sim |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::select(id, time, Cc, dplyr::all_of(group_col))
  dose <- ev |>
    dplyr::filter(evid == 1) |>
    dplyr::select(id, time, amt, dplyr::all_of(group_col))
  fml_conc <- stats::as.formula(paste("Cc ~ time |", group_col, "+ id"))
  fml_dose <- stats::as.formula(paste("amt ~ time |", group_col, "+ id"))
  intervals <- data.frame(
    start = c(0, 120), end = c(24, 144),
    auclast = TRUE, cmax = TRUE, cmin = TRUE
  )
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(as.data.frame(conc), fml_conc),
    PKNCA::PKNCAdose(as.data.frame(dose), fml_dose),
    intervals = intervals
  ))
  as.data.frame(res$result) |>
    dplyr::filter(PPTESTCD %in% c("auclast", "cmax", "cmin")) |>
    dplyr::mutate(day = ifelse(start == 0, "day1", "last"))
}

geo_mean <- function(x) exp(mean(log(x)))
```

```{r table3}
wt_bands <- c(-Inf, 46.5, 57, 70, Inf)
wt_labels <- c("27.6 to 46.5", "46.5 to 57", "57 to 70", "70 to 126")

ev_ado_iv$wt_band <- as.character(cut(ev_ado_iv$WT, wt_bands, labels = wt_labels))
sim_ado_iv$wt_band <- as.character(cut(sim_ado_iv$WT, wt_bands, labels = wt_labels))

nca_ado_iv <- nca_day1_last(sim_ado_iv, ev_ado_iv, "wt_band")

sim_table3 <- nca_ado_iv |>
  dplyr::filter(PPTESTCD %in% c("auclast", "cmax")) |>
  dplyr::mutate(code = paste0(PPTESTCD, "_", day)) |>
  dplyr::group_by(wt_band, code) |>
  dplyr::summarise(value = geo_mean(PPORRES), n = dplyr::n(), .groups = "drop")

sim_table3 |>
  dplyr::distinct(wt_band, n) |>
  dplyr::rename("Body wt (kg)" = wt_band, "Virtual participants" = n) |>
  knitr::kable(caption = "Virtual adolescents per weight band.")

published_table3 <- tibble::tribble(
  ~wt_band,       ~auclast_day1, ~auclast_last, ~cmax_day1, ~cmax_last,
  "27.6 to 46.5", 32.2,          32.4,          3.32,       3.87,
  "46.5 to 57",   29.6,          32.2,          3.83,       3.54,
  "57 to 70",     25.5,          27.9,          2.52,       2.97,
  "70 to 126",    20.1,          22.6,          1.38,       2.29
)

compare_long <- sim_table3 |>
  dplyr::select(wt_band, code, Simulated = value) |>
  dplyr::inner_join(
    tidyr::pivot_longer(
      published_table3, -wt_band,
      names_to = "code", values_to = "Published"
    ),
    by = c("wt_band", "code")
  ) |>
  dplyr::mutate(pct_diff = 100 * (Simulated - Published) / Published)

compare_long |>
  dplyr::mutate(
    Quantity = dplyr::recode(
      code,
      auclast_day1 = "AUC0-24 day 1 (ug*h/mL)",
      auclast_last = "AUC0-24 last day (ug*h/mL)",
      cmax_day1 = "Cmax day 1 (ug/mL)",
      cmax_last = "Cmax last day (ug/mL)"
    )
  ) |>
  dplyr::arrange(code, wt_band) |>
  dplyr::select(Quantity, wt_band, Published, Simulated, pct_diff) |>
  dplyr::rename("Body wt (kg)" = wt_band, "% diff" = pct_diff) |>
  knitr::kable(
    digits = 2,
    caption = paste(
      "Replicates Table 3 of Li 2021: geometric means by body-weight quartile",
      "in adolescents with ABSSSI (intravenous, 200 mg tedizolid phosphate",
      "once daily, last day = day 6)."
    )
  )
```

The eight AUC geometric means reproduce Table 3 to within about 15% (median
absolute difference about 4%). They also show the paper's roughly 30% fall in
exposure from the lightest to the heaviest quartile. The Cmax rows
agree less well. The published day-1 Cmax is non-monotonic across weight (1.38
ug/mL in the heaviest quartile against 3.83 ug/mL in the second), which points
to the actual PN012 regimens (oral doses, switches, sampling) rather than to the
structural model, so Cmax is shown but not asserted.

```{r table3-assert}
auc_rows <- compare_long |> dplyr::filter(grepl("^auclast", code))

# Dosing the prodrug mass instead scales every AUC of this linear model by the
# molecular-weight ratio; the tedizolid basis must fit Table 3 better.
auc_rows$pct_diff_phosphate <- 100 *
  (auc_rows$Simulated * mw_tedizolid_phosphate / mw_tedizolid - auc_rows$Published) /
  auc_rows$Published

stopifnot(
  nrow(auc_rows) == 8,
  # Centre: a mis-transcribed CL, weight exponent or dose unit moves every
  # quartile by tens of percent.
  abs(stats::median(auc_rows$pct_diff)) < 10,
  # Envelope over the eight group geometric means (each from ~50 subjects).
  all(abs(auc_rows$pct_diff) < 25),
  # Exposure falls with body weight, lightest to heaviest quartile.
  auc_rows$Simulated[auc_rows$code == "auclast_last" & auc_rows$wt_band == "27.6 to 46.5"] >
    1.15 * auc_rows$Simulated[auc_rows$code == "auclast_last" & auc_rows$wt_band == "70 to 126"],
  stats::median(abs(auc_rows$pct_diff)) < stats::median(abs(auc_rows$pct_diff_phosphate))
)
```

# Adolescents versus adults (Table 2)

Table 2 compares the 91 PN012 adolescents with 830 adults with ABSSSI from the
earlier phase 2 and 3 trials. The adult trials used oral dosing (PN007, PN009)
or intravenous-to-oral switching (PN010, PN005, PN006), and the route mix of the
830 is not reported. The adult cohort is therefore simulated under both routes,
bracketing the published value. Adult weights are log-normal around 80 kg,
within 40.5-226 kg, the weight range of those trials (supplement Table A2); 8%
have diabetes (supplement Table A1).

```{r table2}
adult_ids <- n_per_arm + seq_len(n_per_arm)
adult_wt <- draw_weights(n_per_arm, 80, 0.22, 40.5, 226)
adult_diab <- stats::rbinom(n_per_arm, 1, 0.08)

ev_adult <- dplyr::bind_rows(
  multi_dose_events(adult_ids, adult_wt, "central", 1L, adult_diab) |>
    dplyr::mutate(group = "Adult, IV"),
  multi_dose_events(adult_ids + n_per_arm, adult_wt, "depot", 1L, adult_diab) |>
    dplyr::mutate(group = "Adult, oral")
)
sim_adult <- as.data.frame(rxode2::rxSolve(
  mod, ev_adult,
  keep = "group", returnType = "data.frame"
))

ev_ado_grp <- ev_ado_iv |> dplyr::mutate(group = "Adolescent, IV")
sim_ado_grp <- sim_ado_iv |> dplyr::mutate(group = "Adolescent, IV")

nca_t2 <- dplyr::bind_rows(
  nca_day1_last(sim_adult, ev_adult, "group"),
  nca_day1_last(sim_ado_grp, ev_ado_grp, "group")
)

sim_table2 <- nca_t2 |>
  dplyr::filter(PPTESTCD %in% c("auclast", "cmax")) |>
  dplyr::mutate(code = paste0(PPTESTCD, "_", day)) |>
  dplyr::group_by(group, code) |>
  dplyr::summarise(value = geo_mean(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = code, values_from = value)

published_table2 <- tibble::tribble(
  ~group,                  ~auclast_day1, ~auclast_last, ~cmax_day1, ~cmax_last,
  "Adult (published)",     22.4,          21.0,          1.81,       2.00,
  "Adolescent (published)", 26.6,         28.6,          2.61,       3.13
)

dplyr::bind_rows(published_table2, sim_table2) |>
  dplyr::rename(
    "Population" = group,
    "AUC0-24 day 1 (ug*h/mL)" = auclast_day1,
    "AUC0-24 last (ug*h/mL)" = auclast_last,
    "Cmax day 1 (ug/mL)" = cmax_day1,
    "Cmax last (ug/mL)" = cmax_last
  ) |>
  knitr::kable(
    digits = 2,
    caption = "Replicates Table 2 of Li 2021 (geometric means)."
  )

adult_iv_auc <- sim_table2$auclast_last[sim_table2$group == "Adult, IV"]
adult_po_auc <- sim_table2$auclast_last[sim_table2$group == "Adult, oral"]
ado_auc <- sim_table2$auclast_last[sim_table2$group == "Adolescent, IV"]

stopifnot(
  # The published adult value lies between the all-oral and all-IV cohorts.
  21.0 > 0.9 * adult_po_auc,
  21.0 < 1.1 * adult_iv_auc,
  abs(100 * (ado_auc - 28.6) / 28.6) < 10,
  # Adolescents are more exposed than adults, as the paper reports.
  ado_auc > adult_iv_auc
)
```

# Concentration-time profiles

```{r profiles, fig.width = 7, fig.height = 4}
sim_ado_po <- as.data.frame(rxode2::rxSolve(
  mod,
  multi_dose_events(ado_ids, ado_wt, "depot", 1L, rep(0L, n_per_arm)),
  returnType = "data.frame"
))

dplyr::bind_rows(
  sim_ado_iv |> dplyr::mutate(admin = "Intravenous"),
  sim_ado_po |> dplyr::mutate(admin = "Oral")
) |>
  dplyr::filter(time >= 120) |>
  dplyr::mutate(tad = time - 120) |>
  dplyr::group_by(admin, tad) |>
  dplyr::summarise(
    p05 = stats::quantile(Cc, 0.05),
    p50 = stats::median(Cc),
    p95 = stats::quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, p50, colour = admin, fill = admin)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.2, colour = NA) +
  geom_line() +
  labs(
    x = "Time after the day-6 dose (h)",
    y = "Tedizolid concentration (ug/mL)",
    colour = NULL, fill = NULL,
    caption = "Median and 5th-95th percentiles, 200 virtual adolescents per route."
  ) +
  theme_bw()
```

# Probability of target attainment (Figure 1)

Figure 1 gives the probability that a 200 mg once-daily adolescent regimen
reaches an fAUC/MIC of at least 3. The text reports the intravenous PTA as 100%
up to an MIC of 0.5 ug/mL, then 96.2%, 40.3% and 1.0% at 1, 2 and 4 ug/mL. The
oral PTA is 99.8%, 90.0%, 24.8% and 0.1% at 0.5, 1, 2 and 4 ug/mL (Results,
"Probability of target attainment"). The paper simulated 1,000 adolescents
drawn from NHANES. Here the same 200 virtual adolescents are used per route,
with the steady-state day-6 AUC(0-24 h).

The Methods give the unbound fraction as 1 - 0.873 = 0.127 (mean protein binding
87.3%), while the Introduction describes tedizolid as about 80% bound. Both are
shown.

```{r pta}
nca_ado_po <- nca_day1_last(
  sim_ado_po |> dplyr::mutate(admin = "Oral"),
  multi_dose_events(ado_ids, ado_wt, "depot", 1L, rep(0L, n_per_arm)) |>
    dplyr::mutate(admin = "Oral"),
  "admin"
)
nca_ado_ivr <- nca_ado_iv |>
  dplyr::mutate(admin = "Intravenous") |>
  dplyr::select(-wt_band)

auc_ss <- dplyr::bind_rows(nca_ado_ivr, nca_ado_po) |>
  dplyr::filter(PPTESTCD == "auclast", day == "last") |>
  dplyr::select(admin, id, auc = PPORRES)

mics <- c(0.5, 1, 2, 4)
pta <- tidyr::expand_grid(
  auc_ss,
  fu = c(0.127, 0.20),
  mic = mics
) |>
  dplyr::group_by(admin, fu, mic) |>
  dplyr::summarise(pta = 100 * mean(fu * auc / mic >= 3), .groups = "drop")

published_pta <- tibble::tribble(
  ~admin,        ~mic, ~published,
  "Intravenous", 0.5,  100,
  "Intravenous", 1,    96.2,
  "Intravenous", 2,    40.3,
  "Intravenous", 4,    1.0,
  "Oral",        0.5,  99.8,
  "Oral",        1,    90.0,
  "Oral",        2,    24.8,
  "Oral",        4,    0.1
)

pta_wide <- pta |>
  dplyr::mutate(fu = paste0("fu_", fu)) |>
  tidyr::pivot_wider(names_from = fu, values_from = pta) |>
  dplyr::left_join(published_pta, by = c("admin", "mic"))

pta_wide |>
  dplyr::rename(
    "Route" = admin,
    "MIC (ug/mL)" = mic,
    "PTA %, fu 0.127" = fu_0.127,
    "PTA %, fu 0.20" = fu_0.2,
    "PTA %, published" = published
  ) |>
  knitr::kable(
    digits = 1,
    caption = "Replicates the PTA values of Figure 1 of Li 2021 (fAUC/MIC >= 3)."
  )
```

```{r pta-plot, fig.width = 7, fig.height = 4}
ggplot(pta, aes(mic, pta, colour = factor(fu))) +
  geom_line() +
  geom_point() +
  geom_point(
    data = published_pta, aes(mic, published),
    inherit.aes = FALSE, shape = 4, size = 3
  ) +
  facet_wrap(~admin) +
  scale_x_log10(breaks = mics) +
  labs(
    x = "MIC (ug/mL)", y = "PTA (%)", colour = "Unbound fraction",
    caption = "Crosses: PTA values reported in the text of Li 2021."
  ) +
  theme_bw()
```

The published curve is reproduced with an unbound fraction of about 0.20
(80% binding) and not with the 0.127 given in the Methods. At 0.127 the PTA at
an MIC of 1 ug/mL falls to about 50-75%, far below the published 90-96%. The
exposures themselves match Table 3, so the gap lies in the PK/PD step and not
in the PK model. Either the figure used the ~80% binding quoted in the
Introduction, or the Methods value does not describe the calculation. The
conclusion the paper draws, 100% PTA up to the 0.5 ug/mL breakpoint, holds
under both values.

```{r pta-assert}
pta_check <- pta_wide |> dplyr::filter(mic %in% c(1, 2))
stopifnot(
  # The ~80%-binding curve tracks the published PTA at the two informative
  # MICs, and the 87.3%-binding curve does not.
  stats::median(abs(pta_check$fu_0.2 - pta_check$published)) < 10,
  stats::median(abs(pta_check$fu_0.127 - pta_check$published)) > 20,
  # PTA at the 0.5 ug/mL breakpoint is essentially 100% under either value.
  all(pta_wide$fu_0.2[pta_wide$mic == 0.5] > 97)
)
```

# Assumptions and deviations

* **Dose basis.** The model is dosed in tedizolid free-base equivalents, so
  200 mg tedizolid phosphate is 164.5 mg. The paper does not state the basis.
  The Table 3 weight-quartile AUCs pick this basis over the prodrug mass, which
  would overpredict every AUC by about 22%.
* **IIV scale.** Table 1 prints IIV as %CV. The variances use
  `omega^2 = log(CV^2 + 1)`, the log-normal relation; for the 258% CV on the
  zero-order duration this gives 2.036, against 6.66 if the CV were read as
  `sqrt(omega^2)`. The paper does not state which convention it used.
* **Residual-error combination.** Table 1 gives the patient-trial and oral-data
  terms as separate fold-increases of the residual SD ("on the square root
  scale"). They are applied multiplicatively, so an oral record from a patient
  trial has SD 0.123 * 4.92 * 2.01 = 1.22 on the log scale. The paper does not
  state whether the two factors combine or one overrides the other. The
  residual error is exponential (`lnorm`), following the log error model of the
  Flanagan 2014 backbone. Residual error does not affect any comparison in this
  article.
* **Study 104 coding.** The phase 2 ABSSSI study 104 shares the phase 3
  residual-error stratum, so `STUDY_PHASE3 = 1` should be set for it.
* **Infection covariate.** The source flag `INFEC` is carried as `DIS_CSSSI`
  (ABSSSI), because the Results name ABSSSI as the covariate and the reference
  was changed to healthy volunteers. How the 41 hospitalized pediatric
  participants with suspected (unconfirmed) infection were coded is not reported.
* **Missing diabetes status.** Status was missing (-99) for 232 participants.
  How that code entered the covariate equation is not reported. The model takes
  `DIS_DIAB` as 0 or 1.
* **Weight exponents.** The Methods say the weight exponents were fixed during
  the update, but Table 1 prints an RSE for each and does not mark them fixed.
  They are carried as estimated values.
* **Infusion time.** Intravenous doses are infused over the modelled duration
  `d2` (0.810 h, fixed, with a fixed 8.26% CV IIV) rather than a duration taken
  from the data set. Dose records must carry `rate = -2`.
* **Bootstrap disagreement.** The bootstrap means for the zero-order duration
  (0.475 h against 0.175 h), `ka` (0.964 against 1.47 1/h) and lag time
  (0.178 against 0.226 h) differ from the final estimates. Only 432 of 1,000
  bootstrap runs were successful, and the bootstrap used FOCE while the final
  model used SAEM. The final-model values are used.
* **Virtual cohorts.** The PN012 weight distribution is log-normal, fitted to the
  median and range of supplement Table A2, with all PN012 participants dosed
  intravenously. The paper's PTA drew adolescents from NHANES; this article
  reuses the PN012-like cohort, which Figure A3 shows to have a comparable
  weight distribution.
* **Exposure-response.** The paper found no exposure-efficacy or
  exposure-safety relationship (Figures 2 and 3) and fitted no model, so there
  is nothing further to encode.
