---
title: "Remimazolam (Liang 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Remimazolam (Liang 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Liang_2026_remimazolam"))$reference`
* Description: `r rxode2::rxode(readModelDb("Liang_2026_remimazolam"))$description`
* Article: <https://doi.org/10.3390/pharmaceutics18030315>
* Supplement (CES1 ontogeny equation, final NONMEM control stream, Tables S1-S5):
  <https://www.mdpi.com/article/10.3390/pharmaceutics18030315/s1>

Remimazolam tosilate is an ultra-short-acting benzodiazepine sedative-hypnotic
acting at GABA-A receptors. Carboxylesterase 1 (CES1) hydrolyses it to the
inactive, renally excreted acid metabolite CNS7054, so clearance is essentially
organ-independent -- the property that motivated the authors to pursue a
paediatric indication in the first place.

The paper pairs two models: the three-compartment population PK model packaged
here, and a whole-body PBPK model built in PK-Sim v12. Both were developed on
adult Phase I data and then extrapolated to children, and the two agreed to
within 1.25-fold on every exposure metric, which is what the authors offer as
mutual cross-validation. Only the PopPK model is reproduced in `nlmixr2lib`;
see *Assumptions and deviations* for why the PBPK model is not.

# Population

The model was fit to pooled data from two single-centre Phase I studies in
healthy Chinese adults (Table 1): **HR7056-Ia**, eleven single ascending dose
groups spanning 0.007 to 0.32 mg/kg given as a 1 min intravenous injection, and
**HR7056-Ib**, a two-period crossover in which every subject received a
0.29 mg/kg loading dose over 1 min followed by a 1.08 mg/kg/h maintenance
infusion for 2 h. Doses are expressed as the free base.

Across both studies 71 subjects contributed 1439 arterial and 518 venous
concentrations. Baseline characteristics (Table S1): 15 female (21.1%) of 71,
mean age 28.6 years (median 27.0, range 18.0-51.0), median weight 62.8 kg
(range 50.2-83.8), median height 168.0 cm, median BMI 22.3 kg/m^2. Hepatic and
renal laboratory values were all within healthy-volunteer ranges.

Two features of the data matter when reusing this model. First, the PopPK model
was fit to the **arterial** concentrations only (Methods 2.2); the venous data
went to the PBPK model. The paper reports marked arterio-venous gradients, with
venous peaks well below the contemporaneous arterial Cmax (Results 3.1), so
this model predicts arterial concentrations and will overpredict a venous
sample taken near the peak. Second, no paediatric data were used at all -- the
3-18-year-old predictions in the paper come from extrapolating this adult model
through its allometric weight term.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Liang_2026_remimazolam")()$population`).

# Source trace

Per-parameter origins are recorded as in-file comments beside each `ini()` entry
in `inst/modeldb/specificDrugs/Liang_2026_remimazolam.R`. Collected here for
review. Note that three entries resolve against the **supplementary NONMEM
control stream** rather than the main text; those are the items the main text
either omits or states differently, and they are discussed under *Assumptions
and deviations*.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Structural model (three compartments, linear CL) | n/a | Results 3.3; `$SUBROUTINES ADVAN11 TRANS4` (Supplement S2) |
| `lcl` (CL) | 1.03 L/min | Table 4, `CL` |
| `lvc` (V1) | 2.08 L | Table 4, `V1` |
| `lvp` (V2) | 10.9 L | Table 4, `V2` |
| `lq` (Q2) | 1.49 L/min | Table 4, `Q2` |
| `lvp2` (V3) | 19.7 L | Table 4, `V3` |
| `lq2` (Q3) | 0.266 L/min | Table 4, `Q3` |
| Allometric scaling form | `(WT / 62)^b` | Equation 3 |
| `wt_ref` | 62 kg | Supplement S2 `$PK` (`WEIGHTBL/62`) and its `$THETA` labels |
| `e_wt_cl`, `e_wt_q`, `e_wt_q2` | 0.75 (fixed) | Equation 3 text; `FIX` on THETA(7), (10), (12) |
| `e_wt_vc`, `e_wt_vp`, `e_wt_vp2` | 1 (fixed) | Equation 3 text; `FIX` on THETA(8), (9), (11) |
| IIV form | exponential | Equation 4 |
| `etalcl` / `etalvc` block | 0.0203; 0.00948; 0.00546 | Table 4 IIV rows; `$OMEGA BLOCK(2)` #1 |
| `etalvp` / `etalq` block | 0.0509; 0.0555; 0.107 | Table 4 IIV rows; `$OMEGA BLOCK(2)` #2 |
| `etalvp2` | 0.072 | Table 4, `omega^2 (V3)` |
| `etalq2` | 0.0777 | Table 4, `omega^2 (Q3)` |
| `expSd` (log-scale residual SD) | `sqrt(0.0162)` | Table 4 `Error sigma^2 (ADD)`; scale from Supplement S2 `$ERROR` / `$SIGMA` |
| `Cc <- 1000 * central / vc` | n/a | Supplement S2 `$PK` (`S1 = V1/1000`); assay in ng/mL (Methods 2.3) |
| Covariate screen (none retained) | n/a | Results 3.2, Table S3 |
| Adult target exposures (validation) | see below | Table 5 |
| Paediatric exposures (validation) | see below | Table 7, PopPK columns |
| Virtual cohort weights | see below | Table S1 (adults), Table S5 (paediatric) |
| CES1 ontogeny (PBPK only, not in this model) | Fbirth 0.20, Age50 1.10 y, n 0.56 | Equation 2, Supplement S1 |

# Virtual cohort

Original observed data are not publicly available. The cohorts below reproduce
the weight distributions the paper used for its own Monte Carlo simulations.

* **Adults** -- normal, mean 63.4 kg, SD 7.3 kg (Methods 2.7.1, matching the
  Table S1 mean and SD).
* **Paediatric, 30 kg or below** -- Table S5 reports median 20.3 kg over
  11.1-30.0 kg; approximated as uniform over that range.
* **Paediatric, above 30 kg** -- Table S5 reports median 46.3 kg over
  30.1-69.8 kg; approximated as uniform over that range.

Each of the six arms uses 200 subjects, the per-arm cap for this repository.
The paper simulated 1000 per arm; the medians and 90% intervals compared below
are stable well below that.

```{r cohort}
# set.seed() seeds R's RNG. It does NOT seed rxode2's simulation RNG, whose
# streams are partitioned per solver thread -- so this cohort is reproducible
# on this machine and different on a machine with a different thread count.
# Every assertion downstream is written to hold for any cohort the model can
# produce (pattern 12 of known-vignette-failure-patterns.md).
set.seed(20260301)

# Dense near the induction peak (Cmax lands at t = 1 min), exactly on t = 120
# for C2h, then progressively coarser through the terminal phase. A grid that
# does not resolve Tmax biases AUC low, so the early phase is sampled at 0.1
# min.
tgrid <- sort(unique(c(
  seq(0, 6, by = 0.1),
  seq(7, 119, by = 1),
  120,
  seq(121, 240, by = 2),
  seq(245, 720, by = 5)
)))

# One arm = n subjects given `ind` mg/kg over 1 min, then `mnt` mg/kg/h for 2 h
# starting when the induction infusion ends. Doses are encoded as amt + rate so
# rxode2 derives the duration (amt / rate): induction 1 min, maintenance 120
# min. `id_offset` keeps subject IDs disjoint across arms -- duplicate IDs are
# silently merged by rxSolve into one subject receiving the summed dose.
make_arm <- function(n, weights, ind, mnt, treatment, id_offset = 0L) {
  subj <- tibble(
    id        = id_offset + seq_len(n),
    WT        = weights,
    treatment = treatment
  )
  induction <- subj |>
    mutate(time = 0, amt = ind * WT, rate = ind * WT, evid = 1L)
  maintenance <- subj |>
    mutate(time = 1, amt = mnt * WT * 2, rate = mnt * WT / 60, evid = 1L)
  obs <- subj |>
    tidyr::crossing(time = tgrid) |>
    mutate(amt = NA_real_, rate = NA_real_, evid = 0L)
  bind_rows(induction, maintenance, obs) |>
    # cmt must name an ODE STATE, never the algebraic observable `Cc`;
    # rxode2 returns Cc as a column at these rows automatically.
    mutate(cmt = "central") |>
    arrange(id, time, desc(evid))
}

n_arm <- 200L
wt_adult <- rnorm(n_arm, mean = 63.4, sd = 7.3)
wt_lo    <- runif(n_arm, min = 11.1, max = 30.0)
wt_hi    <- runif(n_arm, min = 30.1, max = 69.8)

events <- bind_rows(
  make_arm(n_arm, wt_adult, 0.3, 1.0, "Adult 0.3 + 1.0",    id_offset =    0L),
  make_arm(n_arm, wt_adult, 0.3, 3.0, "Adult 0.3 + 3.0",    id_offset =  200L),
  make_arm(n_arm, wt_lo,    0.3, 1.0, "Peds <=30kg 0.3 + 1.0", id_offset =  400L),
  make_arm(n_arm, wt_lo,    0.3, 4.0, "Peds <=30kg 0.3 + 4.0", id_offset =  600L),
  make_arm(n_arm, wt_hi,    0.3, 1.0, "Peds >30kg 0.3 + 1.0",  id_offset =  800L),
  make_arm(n_arm, wt_hi,    0.3, 3.0, "Peds >30kg 0.3 + 3.0",  id_offset = 1000L)
)

# Cheap regression guard against an id_offset mistake.
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))

# Simulated weight medians against the paper's own cohorts.
events |>
  filter(evid == 0) |>
  distinct(treatment, id, WT) |>
  group_by(treatment) |>
  summarise(n = n(), `Median WT (kg)` = round(median(WT), 1), .groups = "drop") |>
  knitr::kable(caption = "Virtual cohorts. Paper medians: 63.4 kg (adult, Methods 2.7.1), 20.3 and 46.3 kg (paediatric, Table S5).")
```

# Simulation

```{r simulate}
mod <- readModelDb("Liang_2026_remimazolam")

# The model declares explicit ODEs rather than linCmt(). Because it also
# exposes a cl / vc pair, rxode2 can auto-convert such a model to its analytic
# solution and discard the ODE system; assert the ODE form survived so a future
# change to that behaviour is caught here rather than silently.
ui <- rxode2::rxode(mod)
stopifnot(is.null(ui$linCmt))
stopifnot(identical(ui$state, c("central", "peripheral1", "peripheral2")))

sim <- rxode2::rxSolve(mod, events = events,
                       keep = c("WT", "treatment")) |>
  as.data.frame()

# Cc must be finite and non-negative everywhere it is defined; a negative
# terminal tail would make PKNCA's aucinf.obs NaN.
obs <- sim |> filter(!is.na(Cc))
stopifnot(nrow(obs) > 0, !anyNA(obs$Cc), all(obs$Cc >= 0))
```

# Replicate published figures

```{r figure-5, fig.width = 7, fig.height = 4.5}
# Replicates Figure 5A of Liang 2026: median and 5th-95th percentile
# concentration-time profiles by body-weight stratum and maintenance rate,
# against the adult target exposure range. The grey band is the adult target
# of Figure 3 / Figure 5: the 5th percentile of 0.3 + 1.0 to the 95th
# percentile of 0.3 + 3.0 (Table 5), i.e. 749 to 3520 ng/mL at 2 h.
adult_band <- tibble(ymin = 749, ymax = 3520)

obs |>
  filter(time <= 240) |>
  group_by(treatment, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_rect(data = adult_band, inherit.aes = FALSE,
            aes(xmin = -Inf, xmax = Inf, ymin = ymin, ymax = ymax),
            fill = "grey70", alpha = 0.35) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~treatment, ncol = 2) +
  scale_y_log10() +
  labs(
    x = "Time (min)", y = "Arterial remimazolam (ng/mL)",
    title = "Concentration-time profiles by weight stratum and maintenance rate",
    caption = "Replicates Figure 5A of Liang 2026 (PopPK panel)."
  )
```

```{r figure-3, fig.width = 6.5, fig.height = 4}
# Replicates Figure 3 of Liang 2026: median profiles at the common
# 0.3 mg/kg + 1.0 mg/kg/h regimen, grouped by the paper's weight bands, showing
# that exposure rises with weight and that the lighter children sit below the
# adult target range.
obs |>
  filter(treatment %in% c("Adult 0.3 + 1.0", "Peds <=30kg 0.3 + 1.0",
                          "Peds >30kg 0.3 + 1.0"),
         time <= 240) |>
  group_by(treatment, time) |>
  summarise(Q50 = quantile(Cc, 0.50), .groups = "drop") |>
  ggplot(aes(time, Q50, colour = treatment)) +
  geom_rect(data = adult_band, inherit.aes = FALSE,
            aes(xmin = -Inf, xmax = Inf, ymin = ymin, ymax = ymax),
            fill = "grey70", alpha = 0.35) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(
    x = "Time (min)", y = "Median arterial remimazolam (ng/mL)",
    colour = NULL,
    title = "Exposure increases with body weight at a fixed mg/kg regimen",
    caption = "Replicates Figure 3 of Liang 2026; grey band is the adult target range (Table 5)."
  ) +
  theme(legend.position = "bottom")
```

# PKNCA validation

Two intervals are needed. `ctrough` over 0-120 min is the paper's C2h (the
concentration 2 h after dosing initiation); `cmax`, `tmax`, `aucinf.obs` and
`half.life` come from the 0 to infinity interval. `ctrough` is `NA` unless a
record sits exactly at the interval end, which is why `tgrid` pins t = 120.

```{r pknca}
# Only `!is.na(Cc)` -- adding `time > 0` or `Cc > 0` would drop the time-zero
# row PKNCA needs to anchor AUC, producing one warning per subject.
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)

# Defensive time-zero guarantee. tgrid already starts at 0 and the infusion
# begins there, so Cc = 0 is the correct pre-dose value; existing rows win.
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "ng/mL", timeu = "min")

# One dose record per infusion; PKNCA needs the total administered amount to
# be consistent with the concentration units.
dose_df <- events |>
  filter(evid == 1) |>
  select(id, time, amt, treatment)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "mg")

intervals <- data.frame(
  start      = c(0, 0),
  end        = c(120, Inf),
  ctrough    = c(TRUE, FALSE),
  cmax       = c(FALSE, TRUE),
  tmax       = c(FALSE, TRUE),
  aucinf.obs = c(FALSE, TRUE),
  half.life  = c(FALSE, TRUE)
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
```

```{r nca-units}
# Model time is minutes, so PKNCA returns AUC in ng*min/mL and half-life in
# min. Convert AUC to the ng*h/mL the paper reports before comparing.
nca_tbl <- as.data.frame(nca_res$result) |>
  mutate(PPORRES = if_else(PPTESTCD == "aucinf.obs", PPORRES / 60, PPORRES))
```

## Comparison against published NCA

Adult reference values are Table 5; paediatric values are the PopPK columns of
Table 7. The paper reports a single Cmax per weight stratum because Cmax is set
by the 0.3 mg/kg induction dose, which is common to both maintenance rates.

```{r nca-compare}
published <- tibble::tribble(
  ~treatment,                ~cmax, ~ctrough, ~aucinf.obs,
  "Adult 0.3 + 1.0",          4990,      946,        2320,
  "Adult 0.3 + 3.0",          4990,     2810,        6360,
  "Peds <=30kg 0.3 + 1.0",    4260,      735,        1760,
  "Peds <=30kg 0.3 + 4.0",    4260,     2910,        6340,
  "Peds >30kg 0.3 + 1.0",     4770,      874,        2130,
  "Peds >30kg 0.3 + 3.0",     4770,     2590,        5830
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_tbl,
  reference     = published,
  by            = "treatment",
  params        = c("cmax", "ctrough", "aucinf.obs"),
  units         = c(cmax = "ng/mL", ctrough = "ng/mL",
                    aucinf.obs = "ng*h/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated vs. published exposures (Table 5 adults, Table 7 PopPK paediatric). Ctrough is the paper's C2h. * differs from reference by >20%.",
  align   = c("l", "l", "r", "r", "r")
)
```

```{r nca-descriptive}
# Tmax and terminal half-life are not reported by the paper; shown for the
# record. Both are in minutes.
nca_tbl |>
  filter(PPTESTCD %in% c("tmax", "half.life")) |>
  group_by(treatment, PPTESTCD) |>
  summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  rename("Regimen" = treatment,
         "Median Tmax (min)" = tmax,
         "Median terminal t1/2 (min)" = "half.life") |>
  knitr::kable(digits = 1, caption = "Simulated Tmax and terminal half-life (not reported in the source).")
```

## Gates

```{r gates}
# ---- Gate 1: deterministic. AUC0-inf of a typical-value solve must equal
# total dose / CL. Both sides use the SAME parameters, so the only difference
# is trapezoidal error on the NCA side; a tight bound is correct here.
mod_typ <- rxode2::zeroRe(mod)
wt_typ  <- 63.4
ev_typ  <- make_arm(1L, wt_typ, 0.3, 1.0, "typical")
sim_typ <- rxode2::rxSolve(mod_typ, events = ev_typ, keep = "WT") |>
  as.data.frame() |>
  filter(!is.na(Cc))
# rxSolve omits `id` entirely for a single-subject event table; PKNCA's
# grouping formula needs it (pattern 8 of known-vignette-failure-patterns.md).
if (is.null(sim_typ$id)) sim_typ$id <- 1L

cl_typ    <- 1.03 * (wt_typ / 62)^0.75          # L/min, Table 4 + Equation 3
dose_typ  <- 0.3 * wt_typ + 1.0 * wt_typ * 2    # mg, induction + 2 h maintenance
auc_exact <- dose_typ * 1000 / cl_typ / 60      # ng*h/mL

auc_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(sim_typ |> select(id, time, Cc), Cc ~ time | id),
  PKNCA::PKNCAdose(data.frame(id = 1L, time = 0, amt = dose_typ),
                   amt ~ time | id),
  intervals = data.frame(start = 0, end = Inf, aucinf.obs = TRUE)
))
auc_nca_h <- as.data.frame(auc_nca$result) |>
  filter(PPTESTCD == "aucinf.obs") |>
  pull(PPORRES) / 60

auc_err_pct <- 100 * (auc_nca_h - auc_exact) / auc_exact
cat(sprintf("closed-form AUC0-inf %.1f vs NCA %.1f ng*h/mL (%.3f%%)\n",
            auc_exact, auc_nca_h, auc_err_pct))
# Pure numerical (trapezoidal) error against the model's own closed form.
stopifnot(abs(auc_err_pct) < 1)

# The same closed form against the paper's printed Table 5 value of 2320
# ng*h/mL. This one number pins the reference weight (62 vs 63 kg), the CL
# estimate, the 0.75 exponent, the 1000x ng/mL scaling and the dose
# interpretation simultaneously.
stopifnot(abs(100 * (auc_exact - 2320) / 2320) < 1)

# ---- Gate 2: cohort medians against the paper. Assert the CENTRE and an
# envelope, never the extremes of a random cohort. Realised across the 18
# comparisons: median |% diff| 1.1 and max 2.7 on one draw, max 2.9 on
# another. Bounds of 10 and 15 leave room for a different cohort draw while
# still going red on a mis-transcribed dose, volume or unit, which move these
# by tens of percent. Do not tighten these to one observed run.
pct <- as.numeric(sub("%", "", cmp$`% diff`))
pct <- pct[!is.na(pct)]
stopifnot(length(pct) == 18L)
cat(sprintf("median |%% diff| %.2f ; max |%% diff| %.2f (n = %d)\n",
            median(abs(pct)), max(abs(pct)), length(pct)))
stopifnot(median(abs(pct)) < 10)
stopifnot(max(abs(pct)) < 15)

# ---- Gate 3: mass balance. Essentially all drug has left the system by the
# end of the grid, so aucinf.obs is not extrapolating a meaningful fraction.
last_row <- sim_typ[which.max(sim_typ$time), ]
frac_left <- with(last_row, central + peripheral1 + peripheral2) / dose_typ
cat(sprintf("fraction of dose remaining at t = %.0f min: %.3g\n",
            last_row$time, frac_left))
stopifnot(frac_left < 0.001)
```

# Assumptions and deviations

**Three items resolve against the supplementary NONMEM control stream rather
than the main text.** The supplement (Section S2) contains the final control
stream, which settles each of them; all three were verified against it.

1. **The residual error is additive on the LOG scale, not on the concentration
   scale.** Equation 5 prints `Cij = CTV,ij + eps_aij` with
   `sigma^2 = 0.0162`, which taken literally is an additive SD of 0.127 ng/mL
   against concentrations reaching 5790 ng/mL -- implausible, and contradicted
   by goodness-of-fit plots drawn on log axes (Figure S3). The control stream
   resolves it: its `$PROBLEM` description reads `logDV+ADD`, `$ERROR` sets
   `IPRED = LOG(F)` and returns `Y = IPRED * (1 + ERR(1)) + ERR(2)`, and
   `$SIGMA` fixes the proportional term to `0 FIX`, leaving only the additive
   term on the log scale. That is nlmixr2's `Cc ~ lnorm(expSd)` with
   `expSd = sqrt(0.0162) = 0.127`, i.e. roughly 12.8% CV -- a sensible assay
   plus model-misspecification magnitude. The inert `0 FIX` proportional term
   is not encoded: it multiplies `log(F)` rather than `F`, so it has no
   `prop()` equivalent, and being fixed at zero it contributes nothing.
2. **The allometric reference weight is 62 kg, not the 63 kg of Table 4's row
   headers.** The control stream centres every term on `(WEIGHTBL/62)` and
   labels its own thetas `L/min/62 kg`; the analysis-population median was
   62.8 kg (Table S1). 62 kg is used here. The choice is not cosmetic but it is
   small -- a 1.3% scale factor on volumes and 1.0% on clearances -- and Gate 1
   above confirms 62 kg is the value that reproduces Table 5's printed
   AUC0-inf of 2320 ng*h/mL to better than 1%.
3. **Q2 carries inter-individual variability, despite Equation 4's prose
   list.** Equation 4 names CL, V1, V2, V3 and Q3 as the parameters with IIV,
   omitting Q2. Table 4 nonetheless reports `omega^2 (Q2) = 0.107` with good
   precision (RSE 18.3%, shrinkage 5%) and a `omega (V2):omega (Q2)`
   covariance, and the control stream declares `ETA(4)` on Q2 inside an
   `$OMEGA BLOCK(2)`. The prose list is simply incomplete; all six disposition
   parameters carry IIV here.

**Further items.**

* **The induction and maintenance infusions are sequential, not concurrent.**
  The paper describes "an induction bolus of 0.3 mg/kg administered over 1 min,
  followed by continuous maintenance infusions", without stating whether the
  maintenance starts at t = 0 or at t = 1 min. Tables 5 and 7 settle it: each
  reports a *single* Cmax shared by both maintenance rates. That can only hold
  if no maintenance drug has been delivered by the time of the peak, so the
  maintenance infusion is encoded here as starting when the induction infusion
  ends. Simulating them concurrently instead raises the 3.0 mg/kg/h arm's Cmax
  about 15% above the 1.0 mg/kg/h arm's and breaks the agreement with Table 5.
* **The companion PBPK model is not packaged.** The paper's PBPK half was built
  in PK-Sim v12 and publishes only a six-row drug-specific parameter table
  (Table 2: molecular weight, logP, pKa, fraction unbound, an optimised
  organ-specific permeability of 4.01e-4 cm/min, and a CES1 specific clearance
  of 50.37 1/min). No ODE system, compartment list, organ volume, blood flow or
  tissue-plasma partition coefficient is reported -- the partition-coefficient
  and cellular-permeability rows read "diverse / PK-Sim standard", naming
  methods rather than values -- and no PK-Sim project file is deposited with
  the supplement. Reconstructing it would require substituting the platform's
  built-in physiology and CES1 tissue-expression libraries from outside any
  on-disk source, which this repository does not do. The PBPK-specific content
  that *is* published is preserved instead: Table 2's parameters and the CES1
  ontogeny function are recorded in the model file's `description` and in the
  `FCES1` entry of `covariatesDataExcluded`.
* **CES1 ontogeny is absent from this model by the authors' own design.** The
  maturation function of Equation 2 (`Fbirth` 0.20, `Adultmax` 1,
  `Age50` 1.10 years, `n` 0.56) was tested on CL and rejected: the analysis
  population was entirely adult and CES1-mature, so FCES1 spanned only
  0.862-0.916 across subjects (Table S1) and changed the OFV negligibly. The
  consequence for reuse is a real limitation the paper acknowledges -- below
  about 3 years of age, where CES1 is genuinely immature, this PopPK model
  scales clearance by weight alone and will overpredict clearance (and so
  underpredict exposure). The paper's own trial enrols from 3 years upward for
  this reason.
* **No covariate other than weight is in the model.** Sex on V2, BUN on CL and
  albumin on V1 all entered the full model during forward inclusion and all
  three were dropped in backward elimination (Table S3). They are documented in
  `covariatesDataExcluded` with their dOFV values so the screen's provenance
  survives; none is referenced in `model()`.
* **Paediatric weight distributions are approximated as uniform.** Table S5
  reports only median [min, max] per stratum. A uniform draw over the reported
  range has expected medians of about 20.6 and 50.0 kg against the paper's
  20.3 and 46.3 kg, so the heavier stratum in particular is simulated a few
  kilograms above the paper's cohort. That accounts for most of the small
  positive bias in the "Peds >30kg" rows of the comparison table.
* **Race is recorded as 100% Asian.** Both studies enrolled Chinese subjects
  only (Table 1). The model carries no race term, so this affects only the
  `population` metadata.
* **Reported minor inconsistencies in the source, none affecting the model.**
  (a) Results 3.1 states 14 females (19.7%) and 57 males (80.3%), while
  Table S1 reports 15 female (21.1%) and 56 male (78.9%); Table S1 is
  internally consistent with its own per-study counts (12 of 63 plus 3 of 8),
  so its values are used. (b) Table 4's "Final Model 95% CI" cell for
  `omega^2 (Q3)` reads 0.0139-0.0267, which is a duplicate of the
  `omega^2 (CL)` row's interval and cannot bracket the 0.0777 point estimate;
  the bootstrap column's 0.0442-0.122 is consistent. Only the point estimate is
  used here, so the erroneous interval has no effect. (c) The tabulated eGFR
  values (median 48.3 mL/min in healthy young adults, Table S1) are not
  reproducible from the MDRD form printed as Equation 6, which evaluates to
  about 124 mL/min for the median subject. eGFR was rejected as a covariate,
  so this affects nothing in the model; it is flagged in the `CRCL` entry of
  `covariatesDataExcluded` in case that column is ever reused.
* **The model predicts arterial concentrations.** See *Population*. A venous
  prediction is not available from this model.
* **Flumazenil co-administration is not modelled.** HR7056-Ib subjects received
  0.5 mg flumazenil or saline 1 h 55 min into dosing. Flumazenil is a receptor
  antagonist with no reported effect on remimazolam disposition, and the source
  model carries no term for it.
