---
title: "Lucitanib (Liao 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Lucitanib (Liao 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Liao_2022_lucitanib"))$reference`
* Description: `r rxode2::rxode(readModelDb("Liao_2022_lucitanib"))$description`
* Article: <https://doi.org/10.1007/s13318-022-00773-w> (open access)
* Supplement: Electronic Supplementary Material 1 of the article (Supplemental
  Table 1, Figures S1-S5 and the NONMEM control stream of the final model).

# Population

Liao 2022 pooled 3540 lucitanib plasma concentrations from 403 adults with
advanced cancers enrolled in five phase 1/2 studies (Table 1): the first-in-human
dose-escalation and expansion study E-3810-I-01 (n = 134, advanced solid
tumours), the phase 2 breast-cancer studies CO-3810-025 (n = 164) and FINESSE
(n = 72), the phase 1b INES study combining lucitanib with fulvestrant (n = 17),
and the phase 2 lung-cancer study E3810-II-02 (n = 16). Lucitanib was given
orally at 5-30 mg once daily, continuously or 15 mg on 5/2 or 21/7 day
schedules, as a film-coated tablet or a hard gelatin capsule. Table 3 gives a
median age of 55 years (range 26-82), a median body weight of 67.5 kg
(35.9-159), 82% women (332/403), and 84% White, 4% Black, 2.5% Asian and 9%
other race. Median creatinine clearance was 95.1 mL/min; 141 patients had mild
and 40 moderate renal impairment, and 80 mild hepatic impairment.

The same information is available programmatically via
`readModelDb("Liao_2022_lucitanib")()$population`.

# Source trace

Every `ini()` value carries its source in an in-file comment in
`inst/modeldb/specificDrugs/Liao_2022_lucitanib.R`. The structural equations
come from the NONMEM control stream of the final model in the Supplemental File
(`$SUBROUTINES ADVAN4 TRANS4`, `$PK`, `$ERROR`).

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL/F, 70 kg) | log(1.90) L/h | Table 4 |
| `lvc` (Vc/F, 70 kg) | log(63.8) L | Table 4 |
| `lka` (ka) | log(4.86) 1/h | Table 4 |
| `lq` (Q/F, 70 kg) | log(6.23) L/h | Table 4 |
| `lvp` (Vp/F, 70 kg) | log(69.1) L | Table 4 |
| `ld1` (D1, capsule) | log(0.814) h | Table 4 |
| `e_form_tablet_d1` | 0.299 | Table 4 'Effect of formulation on D1 (tablet to capsule ratio)'; control stream `TVD1 = THETA(8)*THETA(10)**FORM` |
| `e_wt_cl` | 0.75 (fixed) | Section 2.2; Table 4; control stream `(WT/70)**0.75` on CL and Q |
| `e_wt_vc` | 1 (fixed) | Section 2.2; Table 4; control stream `(WT/70)` on V2 and V3 |
| `etalcl`, `etalvc`, covariance | 0.2058, 0.2952, 0.1469 | Table 4 BSV 47.8% and 58.6%, correlation 0.596; `$OMEGA BLOCK(2)` |
| `etald1` | 0.3084 | Table 4 BSV 60.1% |
| `etalvp` | 0.4627 | Table 4 BSV 76.7% |
| `propSd` | 0.336 | Table 4 'Residual error (proportional, %)'; control stream additive SD `0 FIX` |
| Two-compartment ODEs, depot CMT 1, central CMT 2, peripheral CMT 3 | n/a | Control stream `ADVAN4 TRANS4`; Figure 1 |
| Zero-order release of duration D1 into the depot | n/a | Section 3.2; Figure 1; control stream `D1 = TVD1*EXP(ETA(3))` |
| `Cc = 1000 * central / vc` (ng/mL) | n/a | Control stream `S2 = V2/1000`; Section 2.3 (assay in ng/mL) |

# Typical-value checks

For the 70 kg reference patient the capsule releases its dose into the depot over
D1 = 0.814 h and the tablet over 0.814 x 0.299 = 0.243 h. Every dose record
carries `rate = -2`, which is what makes rxode2 apply the modelled duration
`dur(depot)`; without it the dose is a bolus and D1 has no effect.

```{r typical-setup}
mod <- readModelDb("Liao_2022_lucitanib")
mod_typ <- rxode2::zeroRe(mod)

# Once-daily dosing for n_days days, observations over one chosen interval.
qd_events <- function(dose, n_days, obs_day, form_tablet, wt = 70, step = 0.05) {
  doses <- data.frame(
    time = 24 * (seq_len(n_days) - 1), amt = dose, evid = 1L,
    rate = -2, cmt = "depot"
  )
  obs <- data.frame(
    time = 24 * (obs_day - 1) + seq(0, 24, by = step), amt = 0, evid = 0L,
    rate = 0, cmt = "central"
  )
  ev <- dplyr::bind_rows(doses, obs) |> dplyr::arrange(time, dplyr::desc(evid))
  ev$id <- 1L
  ev$WT <- wt
  ev$FORM_TABLET <- form_tablet
  ev
}

trap <- function(t, y) sum(diff(t) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)

n_days <- 40
typ <- dplyr::bind_rows(lapply(c(0, 1), function(f) {
  s1 <- rxode2::rxSolve(mod_typ, qd_events(15, 1, 1, f), returnType = "data.frame")
  ss <- rxode2::rxSolve(mod_typ, qd_events(15, n_days, n_days, f), returnType = "data.frame")
  prev <- rxode2::rxSolve(mod_typ, qd_events(15, n_days, n_days - 1, f), returnType = "data.frame")
  s1 <- s1[!is.na(s1$Cc), ]
  ss <- ss[!is.na(ss$Cc), ]
  prev <- prev[!is.na(prev$Cc), ]
  data.frame(
    FORM_TABLET = f,
    cmax_d1 = max(s1$Cc), tmax_d1 = s1$time[which.max(s1$Cc)],
    auc_d1 = trap(s1$time, s1$Cc),
    cmax_ss = max(ss$Cc), cmin_ss = min(ss$Cc),
    auc_ss = trap(ss$time, ss$Cc), auc_prev = trap(prev$time, prev$Cc)
  )
}))
typ$Formulation <- ifelse(typ$FORM_TABLET == 1, "Tablet", "Capsule")
```

```{r fig-typical, fig.cap = "Typical-value lucitanib concentrations over the first 15 mg once-daily dosing interval, capsule versus tablet, 70 kg patient."}
dplyr::bind_rows(lapply(c(0, 1), function(f) {
  s <- rxode2::rxSolve(mod_typ, qd_events(15, 1, 1, f), returnType = "data.frame")
  s$Formulation <- ifelse(f == 1, "Tablet (D1 0.243 h)", "Capsule (D1 0.814 h)")
  s[!is.na(s$Cc), ]
})) |>
  ggplot(aes(time, Cc, colour = Formulation)) +
  geom_line(linewidth = 1) +
  labs(x = "Time after first dose (h)", y = "Lucitanib (ng/mL)", colour = NULL) +
  theme_bw()
```

The checks below combine model identities (the steady-state AUC over a dosing
interval must equal dose / (CL/F) regardless of the absorption shape) with
statements in the text of Liao 2022.

```{r typical-checks}
d1_capsule <- 0.814
d1_tablet <- d1_capsule * 0.299
t_half_abs <- log(2) / 4.86
auc_closed <- 1000 * 15 / 1.90 # ng*h/mL; F = 1, CL/F = 1.90 L/h at 70 kg

k <- 1.90 / 63.8
k12 <- 6.23 / 63.8
k21 <- 6.23 / 69.1
s <- k + k12 + k21
beta <- (s - sqrt(s^2 - 4 * k * k21)) / 2
t_half_terminal <- log(2) / beta

cap <- typ[typ$FORM_TABLET == 0, ]
tab <- typ[typ$FORM_TABLET == 1, ]
r_acc <- cap$auc_ss / cap$auc_d1
t_half_eff <- 24 * log(2) / log(r_acc / (r_acc - 1))

checks <- data.frame(
  Quantity = c(
    "Tablet D1 (h)",
    "D1 reduction, tablet vs capsule (%)",
    "Absorption half-life log(2)/ka (h)",
    "AUCtau,ss capsule / (dose / CL/F)",
    "AUCtau,ss tablet / capsule",
    "Cmax day 1, tablet / capsule",
    "Cmax,ss, tablet / capsule",
    "Accumulation ratio AUCtau,ss / AUC0-24 day 1 (capsule)",
    "Effective half-life from accumulation (h)",
    "Terminal half-life log(2)/beta (h)"
  ),
  Model = vapply(signif(c(
    d1_tablet, 100 * (1 - 0.299), t_half_abs,
    cap$auc_ss / auc_closed, tab$auc_ss / cap$auc_ss,
    tab$cmax_d1 / cap$cmax_d1, tab$cmax_ss / cap$cmax_ss,
    r_acc, t_half_eff, t_half_terminal
  ), 4), format, character(1)),
  Paper = c(
    "0.243 (Abstract, Figure 1)", "70 (Discussion)", "0.143 (Discussion)",
    "--", "1 ('AUC would not be influenced', Discussion)",
    "'minimal' (Discussion)", "'minimal' (Discussion)", "--",
    "~31-40 (Discussion, citing an earlier NCA)", "--"
  )
)
knitr::kable(checks, caption = "Typical-value checks, 15 mg once daily, 70 kg patient.")

stopifnot(
  abs(d1_tablet - 0.243) < 5e-4,
  abs(t_half_abs - 0.143) < 5e-4,
  # Steady state reached: successive intervals agree.
  all(abs(typ$auc_ss / typ$auc_prev - 1) < 1e-3),
  # Identity: AUCtau,ss = dose / (CL/F) for both formulations (0.05 h grid).
  all(abs(typ$auc_ss / auc_closed - 1) < 1e-3),
  # The formulation changes only the release duration, so the peak moves by a
  # few percent at most; a 10-fold D1 error would move it by more than 10%.
  abs(tab$cmax_d1 / cap$cmax_d1 - 1) < 0.05,
  abs(tab$cmax_ss / cap$cmax_ss - 1) < 0.05
)
```

The tablet D1, the 70% shorter release and the 0.143 h absorption half-life
reproduce the paper's figures exactly, and the tablet raises the typical Cmax by
about 1% with no change in AUC, the basis of the authors' conclusion that the
formulation effect is not clinically meaningful. The effective half-life from
accumulation (about 42 h) sits just above the ~31-40 h the Discussion quotes from
the earlier non-compartmental analysis of the first-in-human study; the
model's terminal disposition half-life is about 53 h. The paper reports no
model-based half-life, so neither value is gated.

# Virtual cohort

Observed data are not public. The virtual cohort reproduces the dominant
regimens of the pooled data: 10 mg and 15 mg once daily as the capsule,
continuously for 125 days (3000 h, the span of Figure 3). Body weight is
log-normal around the Table 3 median of 67.5 kg with a 20% coefficient of
variation (the paper prints no SD), redrawn when outside the Table 3 range
35.9-159 kg. Sampling follows the first-in-human schedule (Table 1): pre-dose
and 1, 2, 3, 4, 6, 8, 12 and 24 h after the first dose and after the day-28
dose, plus pre-dose samples on days 4, 7, 14 and 21 and every 7 days thereafter.

```{r cohort}
set.seed(20220718)
rxode2::rxSetSeed(20220718)
n_per_arm <- 200
n_days_vpc <- 125

draw_in_range <- function(n, median, cv, lo, hi) {
  x <- stats::rlnorm(n, log(median), sqrt(log(1 + cv^2)))
  bad <- x < lo | x > hi
  while (any(bad)) {
    x[bad] <- stats::rlnorm(sum(bad), log(median), sqrt(log(1 + cv^2)))
    bad <- x < lo | x > hi
  }
  x
}

make_cohort <- function(dose, id_offset) {
  data.frame(
    id = id_offset + seq_len(n_per_arm),
    dose = dose,
    treatment = paste(dose, "mg QD"),
    WT = draw_in_range(n_per_arm, 67.5, 0.20, 35.9, 159),
    FORM_TABLET = 0
  )
}
cohort <- dplyr::bind_rows(make_cohort(10, 0L), make_cohort(15, n_per_arm))
stopifnot(!anyDuplicated(cohort$id))

rich <- c(0, 1, 2, 3, 4, 6, 8, 12, 24)
obs_times <- sort(unique(c(
  rich, 24 * 27 + rich,
  24 * c(3, 6, 13, 20), 24 * seq(34, n_days_vpc - 1, by = 7)
)))

dose_rows <- cohort |>
  dplyr::select(id, dose) |>
  tidyr::crossing(day = seq_len(n_days_vpc)) |>
  dplyr::transmute(id, time = 24 * (day - 1), amt = dose, evid = 1L, rate = -2, cmt = "depot")
obs_rows <- cohort |>
  dplyr::select(id) |>
  tidyr::crossing(time = obs_times) |>
  dplyr::mutate(amt = 0, evid = 0L, rate = 0, cmt = "central")
events <- dplyr::bind_rows(dose_rows, obs_rows) |>
  dplyr::left_join(cohort, by = "id") |>
  dplyr::arrange(id, time, dplyr::desc(evid))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod, events,
  keep = c("treatment", "dose"), returnType = "data.frame"
)
stopifnot(!anyNA(sim$Cc))
```

# Replicate published figures

```{r fig3, fig.cap = "Simulated lucitanib concentrations versus time since first dose (median and 2.5th-97.5th percentiles), 10 and 15 mg once daily as the capsule. Replicates the layout of Figure 3 of Liao 2022, whose prediction intervals pool all studies, doses and sampling times."}
sim |>
  dplyr::group_by(treatment, time) |>
  dplyr::summarise(
    p025 = stats::quantile(Cc, 0.025), p50 = stats::median(Cc),
    p975 = stats::quantile(Cc, 0.975), .groups = "drop"
  ) |>
  ggplot(aes(time, p50, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = p025, ymax = p975), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.8) +
  labs(
    x = "Time since first dose (h)", y = "Plasma lucitanib concentration (ng/mL)",
    colour = NULL, fill = NULL
  ) +
  theme_bw()
```

Figure 3 of Liao 2022 shows a simulated median of roughly 150-300 ng/mL from the
second week onward and a 97.5th percentile of about 1000 ng/mL, over a dataset
dominated by 10 and 15 mg doses sampled at mixed times after dose. The simulated
medians here (trough to peak about 170-330 ng/mL at 10 mg and 250-470 ng/mL at
15 mg on day 28; PKNCA table below) are in that range; the published figure cannot be
compared point by point because it mixes doses and sampling times.

```{r fig-cl, fig.cap = "Individual CL/F versus body weight in the virtual cohort. The line is the typical-value allometric relationship 1.90 x (WT/70)^0.75."}
sim |>
  dplyr::distinct(id, .keep_all = TRUE) |>
  ggplot(aes(WT, cl)) +
  geom_point(alpha = 0.4) +
  geom_function(fun = function(w) 1.90 * (w / 70)^0.75, colour = "steelblue", linewidth = 1) +
  scale_y_log10() +
  labs(x = "Body weight (kg)", y = "Individual CL/F (L/h)") +
  theme_bw()
```

# PKNCA validation

The paper reports no NCA table, so PKNCA characterises the first dosing interval
and the day-28 interval, and checks the steady-state identity AUCtau = dose /
(CL/F) subject by subject with the individual CL/F from the solve.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id, concu = "ng/mL", timeu = "h")
dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id, doseu = "mg")

intervals <- data.frame(
  start = c(0, 24 * 27), end = c(24, 24 * 28),
  cmax = TRUE, tmax = TRUE, auclast = TRUE,
  cmin = c(FALSE, TRUE), cav = c(FALSE, TRUE)
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_tab <- as.data.frame(nca_res)

nca_tab |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "cmin", "cav")) |>
  dplyr::mutate(Interval = ifelse(start == 0, "Day 1", "Day 28")) |>
  dplyr::group_by(treatment, Interval, PPTESTCD) |>
  dplyr::summarise(median = signif(stats::median(PPORRES), 3), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  dplyr::rename(
    Treatment = treatment, `Cmax (ng/mL)` = cmax, `Tmax (h)` = tmax,
    `AUCtau (ng*h/mL)` = auclast, `Cmin (ng/mL)` = cmin, `Cav (ng/mL)` = cav
  ) |>
  knitr::kable(caption = "Median simulated NCA, first dose and day 28 (capsule).")

ind <- sim |>
  dplyr::distinct(id, .keep_all = TRUE) |>
  dplyr::select(id, cl, dose)
idchk <- nca_tab |>
  dplyr::filter(PPTESTCD == "auclast", start == 24 * 27) |>
  dplyr::left_join(ind, by = "id") |>
  dplyr::mutate(ratio = PPORRES / (1000 * dose / cl))
stopifnot(
  nrow(idchk) == 2 * n_per_arm,
  # A 9-point grid puts a few percent of trapezoid error on the narrow peak, and
  # subjects with a large Vp/F are not fully at steady state by day 28; a
  # transcribed CL/F, dose or unit error moves the ratio by tens of percent.
  abs(stats::median(idchk$ratio) - 1) < 0.05,
  stats::quantile(abs(idchk$ratio - 1), 0.9) < 0.15
)
```

The median day-28 AUCtau equals dose / (CL/F) to within a few percent, so the
simulated exposure is consistent with the Table 4 clearance at both dose levels.

# Assumptions and deviations

- **IIV scale.** Table 4 reports between-subject variability as
  "log-proportional, %" without saying whether the percentage is the square root
  of the variance or the coefficient of variation of the log-normal parameter.
  The maintainers converted it as a coefficient of variation, omega^2 =
  log(1 + CV^2), following the package convention. If the percentages are
  instead the square root of the variance, the variances would be 0.228 (CL/F),
  0.343 (Vc/F), 0.361 (D1) and 0.588 (Vp/F); the published bootstrap
  percentile intervals do not distinguish the two readings.
- **Residual error.** The control stream codes a combined error with the
  additive SD fixed to 0, so the model carries the proportional term only.
- **Formulation coding.** The control stream applies THETA(10) (0.299, "tablet
  to capsule ratio") as `THETA(10)**FORM`, and the Abstract gives D1 = 0.243 h
  for the tablet, so `FORM = 1` is the tablet and the capsule is the reference.
  The formulation indicator applies per dose record.
- **Covariates not retained.** The P-gp-inhibitor effect on CL/F and the
  albumin effect on Vc/F were added in forward selection and removed at
  backward elimination; they appear in the final control stream only as
  `1 FIX` multipliers and are omitted here. The formulation effect on F1
  (`1 FIX`) is likewise omitted. Concomitant PPI on ka was statistically
  significant but not included in the final model (Supplemental Table 1 model
  13).
- **Virtual cohort.** Body weight is log-normal around the Table 3 median with
  a 20% coefficient of variation because the paper prints only the median and
  range. Weight is held constant per subject, although the analysis dataset
  coded covariates as time-varying.
- **Correction notices.** None found for Liao 2022 as of 2026-10-05; the only
  lucitanib correction notice in Europe PMC concerns the earlier first-in-human
  report the Discussion cites for the ~31-40 h half-life.
