---
title: "Amikacin (Lim 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Amikacin (Lim 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Lim_2025_amikacin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Lim_2025_amikacin"))$description`
* Article: <https://doi.org/10.1128/aac.00818-25>
* PMC record (open access, with the equation images as supplementary files):
  <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC12326967/>

# Population

Lim 2025 is a single-centre retrospective chart review of every neonate who
received amikacin with therapeutic drug monitoring at the Department of Neonatal
and Developmental Medicine, Singapore General Hospital, between November 2012 and
October 2017. Of 181 eligible neonates, 90 (254 serum concentrations) were used to
build the population PK model and the remaining 91 (280 concentrations) were held
out for external validation.

The model-building cohort (Table 1, "PK modeling" column) was predominantly
extremely and very preterm: median gestational age 29+0 weeks (IQR 26+0 to 31+0,
range 24+0 to 41+0), median birth weight 1.00 kg (IQR 0.87-1.44) and median weight
at amikacin initiation 1.34 kg (IQR 0.90-1.90, range 0.58-3.77). 62.2% were male;
the race distribution was 62.2% Chinese, 13.4% Malay, 12.2% Indian and 12.2%
other. 25.6% were small for gestational age and 11.1% had intrauterine growth
restriction. Serum creatinine at initiation was 43 umol/L (IQR 29-57, range
15-91). No neonate had renal impairment; congenital kidney disease, major
congenital heart disease, acute kidney injury, unstable renal function and ECMO
were exclusion criteria.

The institutional regimen was 11 mg/kg every 36 h below 29 weeks postmenstrual age
and 11 mg/kg every 24 h at or above it, given either as a 30-min intravenous
infusion or as an intramuscular injection. Therapeutic drug monitoring was
performed at the third dose: a trough within 30 min preceding the dose and a peak
60 min after an intramuscular injection or 30 min after the end of a 30-min
intravenous infusion. The institutional targets were 15-25 mg/L for the peak and
< 5 mg/L for the trough.

The same information is available programmatically via
`readModelDb("Lim_2025_amikacin")()$population`.

# Source trace

Lim 2025 contains **no parameter-estimate table**. The entire final model is
reported as two printed equations in Results, "Population PK analysis" (rendered
in the PMC supplementary file set as image `aac.00818-25.m001`):

$$\mathrm{CL\ (L/h)} = 0.0487\,(WT/1{,}000)^{0.919}\,(PMA/30)^{1.53}\,
  \exp\left[\eta \sim N(0,\ \sigma^2 = 0.0188)\right]$$

$$\mathrm{V\ (L/kg)} = 0.497\,(WT/1{,}000)^{0.851}$$

with `WT` in grams (Methods, "Population PK analysis": the weight effect was
explored with WT "expressed in grams") and `PMA` in weeks (same section: the age
covariates are "gestational age (GA [weeks]), PMA (weeks), and PNA (days)").
The reference subject is therefore 1,000 g and 30 weeks postmenstrual age.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL at 1 kg, PMA 30 wk) | `log(0.0487)` L/h | Results, "Population PK analysis", final model equation 1 |
| `e_wt_cl` | 0.919 | Results, final model equation 1 |
| `e_page_cl` | 1.53 | Results, final model equation 1 |
| `lvc` (V at 1 kg) | `log(0.497)` L | Results, "Population PK analysis", final model equation 2 |
| `e_wt_vc` | 0.851 | Results, final model equation 2 |
| `etalcl` (BSV on CL) | 0.0188 (log-scale variance) | Results, final model equation 1, `exp[eta ~ N(0, sigma^2 = 0.0188)]` |
| BSV on V | absent | Results, final model equation 2 carries no `exp[eta]` term |
| `propSd`, `addSd` | `fixed(0)` | Methods / Results state a combined proportional-additive RUV structure; **no values are published anywhere in the paper** |
| One compartment, first-order elimination; `d/dt(central)` | n/a | Results, "Population PK analysis", first sentence; Methods, "Population PK analysis" |
| Dosing into `central` for both IV and IM | n/a | Methods, "Materials and Methods"; the paper reports no absorption-rate parameter for either route |
| Reference WT = 1,000 g, reference PMA = 30 weeks | n/a | The normalising constants printed inside the two equations |

# Reading the two published equations

Two aspects of the printed equations need arithmetic before the model can be used,
and both are worked through here so a reviewer can check them.

## The volume equation is in absolute litres, not L/kg

The V equation is tagged `(L/kg)`, which is the unit named in the Methods sentence
"BSV was evaluated based on V (expressed as liters/kg) and CL (expressed as
liters/hour)". Taken literally it would mean total volume scales as
$WT^{1.851}$, which is not physiologically possible. Two independent checks agree
that the equation returns absolute litres:

```{r v-units}
v_abs   <- function(wt) 0.497 * wt^0.851          # equation returns litres
v_perkg <- function(wt) 0.497 * wt^0.851 * wt     # equation returns L/kg

wt_grid <- c(0.58, 1.34, 3.77)   # min, median, max weight at initiation (Table 1)
v_check <- tibble(
  `Weight (kg)`               = wt_grid,
  `V/WT, litres reading (L/kg)` = v_abs(wt_grid) / wt_grid,
  `V/WT, L/kg reading (L/kg)`   = v_perkg(wt_grid) / wt_grid
)
knitr::kable(v_check, digits = 3,
             caption = "Weight-normalised volume of distribution under the two readings.")
```

Amikacin is a hydrophilic aminoglycoside confined to extracellular fluid, so its
weight-normalised volume in neonates is roughly 0.4-0.6 L/kg and **falls** as the
extracellular fluid fraction falls with maturation. The litres reading gives
0.54 to 0.41 L/kg across the cohort's weight range, in the right band and falling.
The L/kg reading gives 0.31 to 1.54 L/kg, rising, and reaches three times the
plausible value at the heaviest weight.

The paper's own Table 3 settles it independently. A single 11 mg/kg dose gives an
immediate post-dose concentration of `dose / V`:

```{r v-peak-check}
c0 <- function(wt, v) 11 * wt / v(wt)
tibble(
  Reading = c("V in litres", "V in L/kg"),
  `C0 at the median 1.34 kg (mg/L)` = c(c0(1.34, v_abs), c0(1.34, v_perkg))
) |>
  knitr::kable(digits = 2,
               caption = "Predicted immediate post-dose concentration versus the observed peaks of Table 3 (IV 22.23 mg/L, IM 24.46 mg/L).")
```

The litres reading lands on the observed peaks; the L/kg reading is roughly 25%
low. The model file therefore encodes `vc <- exp(lvc) * (WT / 1)^e_wt_vc` in
litres, and records the unit discrepancy in its `ini()` comments.

## Table 1's postmenstrual-age column carries a unit error

Table 1 prints "PMA at initiation" as median 45+6 weeks (IQR 39+5 to 53+5, range
25+6 to 113+6). That column is inconsistent with the rest of the paper, and the
inconsistency is exactly what you get from computing postmenstrual age as
`GA_weeks + PNA_days` instead of `GA_weeks + PNA_days / 7`:

```{r pma-reconstruction}
pma_recon <- function(ga, pma_printed) ga + (pma_printed - ga) / 7

pma_tbl <- tibble(
  Statistic          = c("Q1", "Median", "Q3"),
  `GA (weeks)`       = c(26 + 0/7, 29 + 0/7, 31 + 0/7),
  `PMA printed (weeks)` = c(39 + 5/7, 45 + 6/7, 53 + 5/7)
) |>
  mutate(
    `Implied PNA (days)`     = `PMA printed (weeks)` - `GA (weeks)`,
    `PMA reconstructed (weeks)` = pma_recon(`GA (weeks)`, `PMA printed (weeks)`)
  )
knitr::kable(pma_tbl, digits = 2,
             caption = "Table 1's GA and PMA columns, and the postmenstrual age implied by reading the residual as days rather than weeks.")
```

Four independent checks agree that the reconstructed column, median about 31
weeks, is the true one.

1. **Growth.** Median birth weight is 1.00 kg and median weight at initiation is
   1.34 kg. The printed PMA implies 16.9 **weeks** of postnatal life, i.e. 20 g
   of weight gain per week -- roughly a seventh of normal preterm growth. The
   reconstruction implies 16.9 **days**, i.e. about 140 g per week, which is the
   expected 8-15 g/kg/day.
2. **The dosing rule.** The regimen splits at 29 weeks postmenstrual age. Under
   the printed column the first quartile is already 39.7 weeks, so the q36h arm
   would be empty. Under the reconstruction the quartiles straddle 29 weeks and
   the rule partitions the cohort.
3. **Trough concentrations.** Shown below: the model driven by the printed PMA
   under-predicts Table 3's observed troughs about sevenfold, while the
   reconstruction reproduces them.
4. **Internal consistency of Table 1 itself.** The printed PMA maximum of 113+6
   weeks is 2.2 years, which no reading of "neonates" supports; as days it is a
   90-day admission, which the cohort description does support.

This affects only what postmenstrual ages a user should regard as representative
of the source cohort. **It does not affect the model equations**, which are
unambiguous. The virtual cohort below uses the reconstruction, and the deviation
is recorded in the model file's `covariateData$PAGE` notes and in
"Assumptions and deviations".

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces the
Table 1 marginal distributions (weight at initiation, gestational age) and the
reconstructed postmenstrual age, and applies the institutional regimen, with each
subject assigned to the q24h or q36h arm by the paper's own 29-week rule. The
covariates are drawn from R's RNG and are therefore reproducible; only the
between-subject random effect on clearance is drawn by rxode2.

```{r cohort}
# set.seed() seeds R's RNG, which governs the covariate draws below. It does NOT
# seed rxode2's simulation RNG, whose streams are partitioned per solver thread,
# so the eta draws differ between a 2-core CI runner and a 16-thread
# workstation. Every assertion downstream is written to hold for any cohort the
# model can produce; see pattern 12 of known-vignette-failure-patterns.md.
set.seed(20250717)

n_total <- 250L   # splits into ~184 q24h and ~66 q36h; both arms stay under the 200/arm cap

qn75 <- stats::qnorm(0.75)

# Weight at initiation: median 1.34 kg, IQR 0.90-1.90, range 0.58-3.77 (Table 1).
wt_sdlog <- log(1.90 / 0.90) / (2 * qn75)
WT <- pmin(pmax(rlnorm(n_total, log(1.34), wt_sdlog), 0.58), 3.77)

# Gestational age: median 29+0 wk, IQR 26+0 to 31+0, range 24+0 to 41+0 (Table 1).
ga_sd <- (31 - 26) / (2 * qn75)
GA <- pmin(pmax(rnorm(n_total, 29, ga_sd), 24), 41)

# Postnatal age in days, from the reconstruction above: median 16.86 d,
# IQR 13.71-22.71 d.
pna_sdlog <- log(22.71 / 13.71) / (2 * qn75)
PNA_days <- rlnorm(n_total, log(16.857), pna_sdlog)

PAGE <- GA + PNA_days / 7

subj <- tibble(
  id        = seq_len(n_total),
  WT        = WT,
  GA        = GA,
  PAGE      = PAGE,
  tau       = ifelse(PAGE < 29, 36, 24),
  treatment = ifelse(PAGE < 29, "11 mg/kg q36h (PMA < 29 wk)",
                                "11 mg/kg q24h (PMA >= 29 wk)"),
  amt       = 11 * WT
)

stopifnot(all(table(subj$treatment) <= 200))   # 200-per-arm cap

knitr::kable(
  subj |>
    group_by(treatment) |>
    summarise(
      n = n(),
      `WT median (kg)`   = median(WT),
      `WT IQR (kg)`      = sprintf("%.2f-%.2f", quantile(WT, 0.25), quantile(WT, 0.75)),
      `PMA median (wk)`  = median(PAGE),
      `PMA IQR (wk)`     = sprintf("%.1f-%.1f", quantile(PAGE, 0.25), quantile(PAGE, 0.75)),
      .groups = "drop"
    ),
  digits = 2,
  caption = "Virtual cohort by dosing arm."
)
```

```{r events}
# Three doses (TDM is performed at the third), each a 30-min infusion.
dosing <- subj |>
  tidyr::crossing(occ = 0:2) |>
  mutate(time = occ * tau, evid = 1L, cmt = "central", dur = 0.5) |>
  select(id, WT, PAGE, treatment, tau, time, amt, evid, cmt, dur)

# Dense grid over the three intervals, for the profile figure and for NCA.
grid_obs <- subj |>
  select(id, WT, PAGE, treatment, tau) |>
  tidyr::crossing(frac = seq(0, 3, by = 1 / 48)) |>
  mutate(time = frac * tau) |>
  select(id, WT, PAGE, treatment, tau, time)

# The protocol TDM samples: trough immediately before the third dose, peak 1 h
# after that infusion starts (= 30 min after the end of a 30-min infusion).
tdm_obs <- subj |>
  select(id, WT, PAGE, treatment, tau) |>
  tidyr::crossing(off = c(-0.1, 1.0)) |>
  mutate(time = 2 * tau + off) |>
  select(id, WT, PAGE, treatment, tau, time)

obs <- bind_rows(grid_obs, tdm_obs) |>
  distinct(id, time, .keep_all = TRUE) |>          # the grid can land on a TDM time
  mutate(amt = NA_real_, evid = 0L, cmt = "central", dur = NA_real_)

events <- bind_rows(dosing, obs) |> arrange(id, time, desc(evid))
stopifnot(nrow(distinct(events, id, time, evid)) == nrow(events))
```

# Simulation

```{r simulate}
mod <- readModelDb("Lim_2025_amikacin")

sim <- rxode2::rxSolve(mod, events = events,
                       keep = c("treatment", "tau")) |>
  as.data.frame()

stopifnot(!anyNA(sim$Cc), all(sim$Cc >= 0))
```

## The solve reproduces its own closed form

The model is a one-compartment system with an infusion, so its solution is known
exactly. This is a deterministic check between two expressions of the same
parameters, so it is asserted tightly.

```{r closed-form}
mod_typ <- rxode2::zeroRe(mod)
sim_typ <- rxode2::rxSolve(mod_typ, events = events,
                           keep = c("treatment", "tau")) |>
  as.data.frame()

cl_of <- function(wt, pma) 0.0487 * wt^0.919 * (pma / 30)^1.53
v_of  <- function(wt)      0.497  * wt^0.851

# Superposition of three 30-min infusions of rate amt/0.5.
cc_closed <- function(t, wt, pma, tau) {
  k <- cl_of(wt, pma) / v_of(wt)
  R <- 11 * wt / 0.5
  one <- function(dt) {
    ifelse(dt <= 0, 0,
      ifelse(dt <= 0.5,
             R / (k * v_of(wt)) * (1 - exp(-k * dt)),
             R / (k * v_of(wt)) * (1 - exp(-k * 0.5)) * exp(-k * (dt - 0.5))))
  }
  one(t) + one(t - tau) + one(t - 2 * tau)
}

chk <- sim_typ |>
  filter(time > 0) |>
  mutate(closed = cc_closed(time, WT, PAGE, tau),
         rel = abs(Cc - closed) / closed)

stopifnot(max(chk$rel) < 1e-6)
sprintf("max relative difference, ODE solve vs closed form: %.2e", max(chk$rel))
```

# Replicate published figures

```{r figure-3}
# Companion to Figure 3 of Lim 2025 (the pcVPC of the final model): the
# concentration-time profile over the three doses that precede TDM, by arm.
sim |>
  group_by(treatment, tafd = time / tau) |>
  summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tafd, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~treatment) +
  labs(x = "Dosing intervals since first dose",
       y = "Amikacin serum concentration (mg/L)",
       title = "Simulated amikacin profiles over the first three doses",
       caption = "Companion to Figure 3 of Lim 2025. Band is the 5th-95th percentile.")
```

```{r figure-5}
# Replicates Figure 5 of Lim 2025: the TDM peak and trough concentrations at the
# third dose against the institutional target ranges (peak 15-25 mg/L,
# trough < 5 mg/L).
tdm <- sim |>
  mutate(sample = case_when(
    abs(time - (2 * tau - 0.1)) < 1e-8 ~ "Trough (pre-3rd dose)",
    abs(time - (2 * tau + 1.0)) < 1e-8 ~ "Peak (1 h after 3rd dose start)",
    TRUE ~ NA_character_
  )) |>
  filter(!is.na(sample))

stopifnot(nrow(tdm) == 2L * n_total)   # exactly one peak and one trough per subject

targets <- tibble(
  sample = c("Peak (1 h after 3rd dose start)", "Peak (1 h after 3rd dose start)",
             "Trough (pre-3rd dose)"),
  y      = c(15, 25, 5)
)

ggplot(tdm, aes(treatment, Cc)) +
  geom_boxplot(outlier.alpha = 0.3) +
  geom_hline(data = targets, aes(yintercept = y), linetype = "dashed",
             colour = "firebrick") +
  facet_wrap(~sample, scales = "free_y") +
  coord_flip() +
  labs(x = NULL, y = "Amikacin serum concentration (mg/L)",
       title = "Third-dose TDM concentrations against the institutional targets",
       caption = "Replicates Figure 5 of Lim 2025. Dashed lines are the 15-25 mg/L peak and < 5 mg/L trough targets.")
```

```{r target-attainment}
attain <- tdm |>
  group_by(sample) |>
  summarise(
    `Within target (%)` = 100 * mean(
      if (first(sample) == "Trough (pre-3rd dose)") Cc < 5 else Cc >= 15 & Cc <= 25
    ),
    .groups = "drop"
  ) |>
  mutate(`Lim 2025 observed (%)` = c(61, 99))

knitr::kable(attain, digits = 1,
             caption = "Target attainment at the third dose, simulated vs. Lim 2025 Results.")
```

The trough result reproduces the paper (99% observed, effectively all simulated
subjects below 5 mg/L). The **peak** attainment is higher in the simulation than
the 61% Lim 2025 observed, and this is expected rather than a discrepancy in the
structural model: the packaged model carries between-subject variability on
clearance only, because that is the only random effect the paper published. The
residual error structure is reported but its magnitude is not, and there is no
published between-subject variability on volume -- yet peak concentrations are
governed almost entirely by volume and by assay and sample-timing error. The
simulated peaks are therefore much less dispersed than the observed ones, so a
larger fraction falls inside a fixed 10 mg/L-wide window. The **central tendency**
is what this model can be held to, and it is checked below.

# PKNCA validation

NCA is run over the third dosing interval, which is the interval in which
therapeutic drug monitoring was performed, separately for each dosing arm.

```{r pknca}
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  filter(time >= 2 * tau, time <= 3 * tau) |>
  mutate(time_int = time - 2 * tau) |>
  select(id, time = time_int, Cc, treatment)

# Guarantee a time-zero (start-of-interval) record for every subject.
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- subj |>
  mutate(time = 0) |>
  select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start     = 0,
  end       = c(24, 36),
  treatment = c("11 mg/kg q24h (PMA >= 29 wk)", "11 mg/kg q36h (PMA < 29 wk)"),
  cmax      = TRUE,
  tmax      = TRUE,
  auclast   = TRUE,
  half.life = TRUE,
  cmin      = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_summary <- as.data.frame(nca_res$result) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "half.life", "cmin")) |>
  group_by(treatment, PPTESTCD) |>
  summarise(median = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median)

nca_summary |>
  select(treatment, cmax, tmax, auclast, half.life, cmin) |>
  rename(
    "Dosing arm"        = treatment,
    "Cmax (mg/L)"       = cmax,
    "Tmax (h)"          = tmax,
    "AUClast (mg*h/L)"  = auclast,
    "t1/2 (h)"          = half.life,
    "Cmin (mg/L)"       = cmin
  ) |>
  knitr::kable(digits = 2,
               caption = "Median NCA parameters over the third dosing interval, by arm.")
```

The median terminal half-life is around 6-7 h, consistent with the paper's
reasoning that "most of the amikacin concentrations were measured after the
distribution phase" so a one-compartment model sufficed, and with the extended
36 h interval being used for the least mature neonates.

```{r halflife-gate}
hl <- as.data.frame(nca_res$result) |> filter(PPTESTCD == "half.life")
# Deterministic band: the packaged model's t1/2 = ln(2) * V/CL is set entirely by
# WT and PMA. The cohort spans t1/2 from roughly 4 to 11 h; the median cannot
# move outside 4-10 h unless a structural constant was mis-transcribed.
stopifnot(median(hl$PPORRES) > 4, median(hl$PPORRES) < 10)
```

# Comparison against published values

Lim 2025 reports no NCA table. It does report, in Table 3, the mean peak and
trough serum concentrations actually measured at the third dose, separately for
the intravenous and intramuscular routes. Those are the paper's only published
exposure numbers and they are directly comparable to the protocol-matched samples
drawn from the simulation above.

The packaged model has **no route term** -- Lim 2025 modelled both routes with the
same one-compartment structure and reported no absorption-rate parameter -- so the
simulated column is identical for the two reference rows, by construction.

```{r nca-compare}
sim_tdm <- tdm |>
  group_by(sample) |>
  summarise(value = median(Cc), .groups = "drop") |>
  mutate(PPTESTCD = ifelse(grepl("^Peak", sample), "peak_tdm", "trough_tdm"))

simulated_tbl <- tidyr::crossing(
  Route = c("Intravenous", "Intramuscular"),
  sim_tdm |> select(PPTESTCD, PPORRES = value)
)

published <- tibble::tribble(
  ~Route,          ~peak_tdm, ~trough_tdm,
  "Intravenous",   22.23,     1.62,
  "Intramuscular", 24.46,     1.65
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = simulated_tbl,
  reference     = published,
  by            = "Route",
  units         = c(peak_tdm = "mg/L", trough_tdm = "mg/L"),
  tolerance_pct = 20,
  label_first_column = "TDM concentration"
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated third-dose TDM concentrations vs. Lim 2025 Table 3 observed means.",
    "peak_tdm is sampled 1 h after the start of the 30-min infusion and",
    "trough_tdm immediately before the third dose, matching the paper's TDM",
    "definitions. * differs from reference by more than 20%."
  ),
  align = c("l", "l", "r", "r", "r")
)
```

Every cell is within 20% of the published mean. The simulated peak sits between
the observed intravenous and intramuscular means and the simulated trough matches
both within a few percent.

```{r tdm-gate}
peak_med   <- median(tdm$Cc[grepl("^Peak",   tdm$sample)])
trough_med <- median(tdm$Cc[grepl("^Trough", tdm$sample)])

# Structural gates. Both quantities are dominated by the covariate model and the
# 11 mg/kg dose; the only stochastic input is the published BSV on CL
# (variance 0.0188, i.e. about 14% CV), so these bounds carry ample headroom
# over cohort-to-cohort noise while still failing hard on a mis-transcribed
# volume, dose, exponent or reference value -- each of which moves these
# numbers by tens of percent. Realised on this render: peak 22.7, trough 1.5.
stopifnot(abs(peak_med   - 22.23) / 22.23 < 0.15)   # Table 3 IV mean
stopifnot(trough_med > 0.7, trough_med < 3.5)       # Table 3 means 1.62 / 1.65

# Essentially every subject should be below the 5 mg/L trough target, as in the
# paper's 99%.
stopifnot(mean(tdm$Cc[grepl("^Trough", tdm$sample)] < 5) > 0.90)
```

## The printed postmenstrual ages do not reproduce Table 3

This is the third of the four checks listed earlier. Re-running the same typical-value
cohort with Table 1's printed PMA column, instead of the reconstruction, moves the
predicted trough far away from the observed values:

```{r pma-falsifier}
# Same subjects, same weights, same regimen -- only PMA changes. Printed PMA is
# recovered by inverting the reconstruction: PMA_printed = GA + PNA_days.
subj_printed <- subj |> mutate(PAGE = GA + (PAGE - GA) * 7)

events_printed <- events |>
  select(-PAGE) |>
  left_join(subj_printed |> select(id, PAGE), by = "id")

trough_printed <- rxode2::rxSolve(mod_typ, events = events_printed,
                                  keep = c("treatment", "tau")) |>
  as.data.frame() |>
  filter(abs(time - (2 * tau - 0.1)) < 1e-8)

trough_recon <- sim_typ |> filter(abs(time - (2 * tau - 0.1)) < 1e-8)

falsifier <- tibble(
  `PMA column`            = c("Table 1 as printed", "Reconstructed (GA + PNA/7)"),
  `Median PMA (weeks)`    = c(median(subj_printed$PAGE), median(subj$PAGE)),
  `Median trough (mg/L)`  = c(median(trough_printed$Cc), median(trough_recon$Cc)),
  `Lim 2025 Table 3 (mg/L)` = c(1.62, 1.62)
)
knitr::kable(falsifier, digits = 2,
             caption = "Typical-value trough at the third dose under the two readings of Table 1's PMA column.")

# Deterministic (typical-value) comparison: the printed column under-predicts the
# observed trough by roughly sevenfold, the reconstruction lands on it.
stopifnot(median(trough_printed$Cc) < 0.6)
stopifnot(abs(median(trough_recon$Cc) - 1.62) / 1.62 < 0.5)
```

# Assumptions and deviations

- **The V equation's `(L/kg)` unit tag is not used.** The printed equation is
  tagged L/kg but is encoded as absolute litres, for the two reasons worked
  through in "Reading the two published equations": the L/kg reading implies total
  volume scaling as $WT^{1.851}$ and a weight-normalised volume of 1.54 L/kg at
  the cohort's heaviest weight (roughly three times the value amikacin's
  extracellular distribution allows, and rising rather than falling with
  maturation), and it under-predicts Table 3's observed peaks by about 25%. The
  litres reading is physiologically correct on both counts. The `(L/kg)` tag
  matches the Methods sentence describing how BSV was parameterised during model
  building, and appears to have been carried onto the final equation.

- **Table 1's "PMA at initiation" column is not used for the virtual cohort.**
  It is inconsistent with Table 1's own GA and weight columns, with the paper's
  own 29-week dosing rule, and with Table 3's observed troughs, in the specific
  way produced by computing `GA_weeks + PNA_days` rather than
  `GA_weeks + PNA_days / 7`. The cohort uses the reconstruction (median about 31
  weeks). This is a **deviation from a printed table**, not a change to the model:
  the model equations are used exactly as published, and a user supplying true
  postmenstrual ages gets the published behaviour. The reconstruction is recorded
  in `covariateData$PAGE$notes` and in the `population$page_range` field so it
  cannot be lost.

- **No between-subject variability on volume.** The published V equation carries
  no `exp[eta]` term while the CL equation printed immediately above it does, so
  the omission is treated as informative rather than as a reporting gap. It is
  corroborated by the Discussion ("there was a good correlation between body WT
  and volume of distribution (V), which matches our expectation"). No eta is
  declared on `lvc`.

- **Residual error is encoded at `fixed(0)`.** Lim 2025 states that a combined
  proportional-additive residual error model was selected, but publishes neither
  coefficient: the paper has no parameter-estimate table, and the PMC
  supplementary file set contains only figure and equation images, no control
  stream. Both `propSd` and `addSd` are therefore `fixed(0)` rather than invented,
  with the combined structure retained so that a user who obtains the estimates
  can substitute them directly. Consequently `sim` equals `ipred` in any solve of
  this model, and simulated concentrations are less dispersed than real
  observations -- this is the reason the simulated peak-target attainment (about
  87%) exceeds the paper's observed 61%, as discussed above.

- **No uncertainty on any parameter.** The paper reports a 2,000-replicate
  bootstrap as "satisfactory" but publishes no standard errors, relative standard
  errors, or confidence intervals for any estimate, so none are recorded.

- **Route of administration is not modelled.** Amikacin was given intravenously
  or intramuscularly, and Table 3 compares the two, but the paper models both with
  the same one-compartment structure and reports no absorption-rate or
  bioavailability parameter for the intramuscular route. Doses of either route
  enter `central` directly. A user wanting to distinguish the routes would need to
  add an absorption compartment, which this paper does not parameterise.

- **The virtual cohort's covariate distributions are assumed.** Table 1 reports
  medians, interquartile ranges and extremes but not distributional forms. Weight
  at initiation and postnatal age are drawn as log-normal and gestational age as
  normal, each matched to the reported median and IQR and truncated to the
  reported range. Weight is held constant per subject over the three simulated
  doses, whereas the source data set carried it as time-varying with
  last-observation-carried-forward imputation; over the 48-72 h simulated here the
  difference is negligible.

- **Serum creatinine is not in the model.** It was screened as a power covariate
  on CL and not retained; the paper's Discussion notes that renal maturation in
  neonates is captured by postmenstrual age instead. No `SCR` column is required.
