---
title: "Recombinant Erwinia chrysanthemi asparaginase (Lin 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recombinant Erwinia chrysanthemi asparaginase (Lin 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r model}
mod <- rxode2::rxode(readModelDb("Lin_2023_asparaginaseErwiniaRecombinant"))
```

# Model and source

* Citation: `r mod$reference`
* Description: `r mod$description`
* Article: <https://doi.org/10.1111/cts.13499>
* Upstream phase I population PK model that supplied the three fixed
  absorption parameters: <https://doi.org/10.1002/cpdd.1002>
* Upstream phase I study reporting the noncompartmental analysis, cited as
  reference 16 of Lin 2023 and *not* used as a source here: Lin T,
  Hernandez-Illas M, Rey A, et al. *Clin Transl Sci.* 2021;14:870-879.

JZP458 (marketed as Rylaze) is a recombinant *Erwinia chrysanthemi*
asparaginase produced in a *Pseudomonas fluorescens* expression platform. It is
given to patients with acute lymphoblastic leukemia (ALL) or lymphoblastic
lymphoma (LBL) who have developed hypersensitivity to *E. coli*-derived
asparaginases. The measured quantity is **serum asparaginase activity (SAA)**,
and the therapeutic target is a nadir SAA (NSAA) of at least 0.1 IU/mL, the
accepted threshold for adequate asparagine depletion.

# Population

The model was fit to 2687 SAA observations (2145 quantifiable, 542 below the
limit of quantitation, handled by the Beal M3 method) from 166 patients in the
phase II/III AALL1931 study (NCT04145531), run with the Children's Oncology
Group. Patients were 1.7 to 25 years old (mean 10.2, median 10.0 years), weighed
9.33 to 131.0 kg (median 36.5 kg), and had a body surface area of 0.44 to
2.53 m^2 (mean 1.23, median 1.17 m^2). 38.0% were female. Race was 68.7%
White/Caucasian, 13.3% Black/African American, 4.2% Asian, 1.8% American
Indian/Alaska Native, 0.6% Other and 11.4% declined to state; 31.3% were
Hispanic or Latino. Primary disease was ALL in 89.8% and LBL in 10.2%, with
disease subtype B-ALL 74.1%, T-ALL 15.7%, T-LBL 9.6% and B-LBL 0.6%. All 166
patients had received pegaspargase before study entry, and exactly half were
antidrug-antibody positive. Three intramuscular Monday/Wednesday/Friday cohorts
were enrolled: 25 mg/m^2 (n = 32), 37.5 mg/m^2 (n = 83) and 25/25/50 mg/m^2
(n = 51). Baseline demographics are Lin 2023 Table 1.

The same information is available programmatically from the model's
`population` metadata:

```{r population}
str(mod$population, max.level = 1)
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Lin_2023_asparaginaseErwiniaRecombinant.R` carries
an in-file comment naming its origin. They are collected here for review.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` (CL) | 146 mL/h | Lin 2023 Table 2 (SE 4.93, RSE 3.38%, 95% CI 136-156) |
| `lvc` (V) | 445 mL | Lin 2023 Table 2 (SE 34.9, RSE 7.85%, 95% CI 377-513) |
| `lka` (Ka) | 0.0369 1/h, fixed | Lin 2023 Table 2 (no SE/RSE/CI); fixed from the phase I model of Lin 2021 |
| `lr1` (R1) | 1810 IU/h, fixed | Lin 2023 Table 2 (no SE/RSE/CI); fixed from the phase I model of Lin 2021 |
| `lfdepot` (F) | 0.359, fixed | Lin 2023 Table 2 (no SE/RSE/CI); fixed from the phase I model of Lin 2021 |
| `e_bsa_cl` | 1.48 | Lin 2023 Table 2 (SE 0.0686, RSE 4.63%, 95% CI 1.35-1.61) |
| `e_bsa_vc` | 1.61 | Lin 2023 Table 2 (SE 0.151, RSE 9.4%, 95% CI 1.31-1.91) |
| `e_black_cl` | 0.674 | Lin 2023 Table 2 (SE 0.0715, RSE 10.6%, 95% CI 0.534-0.814) |
| `e_tall_cl` | 0.771 | Lin 2023 Table 2 (SE 0.0686, RSE 8.89%, 95% CI 0.637-0.905) |
| `etalcl` | 0.125 (CV 36.5%) | Lin 2023 Table 2 (SE 0.0173, RSE 13.8%, 95% CI 0.0911-0.159) |
| `etalvc` | 0.479 (CV 78.4%) | Lin 2023 Table 2 (SE 0.0867, RSE 18.1%, 95% CI 0.309-0.649) |
| `propSd` | 0.46 | Lin 2023 Table 2 (SE 0.00665, RSE 1.45%, 95% CI 0.447-0.473) |
| `addSd` | 0.0166 IU/mL | Lin 2023 Table 2 (SE 0.000539, RSE 3.25%, 95% CI 0.0155-0.0177) |
| CL covariate equation | `146 * (BSA/1.2)^1.48 * 0.674^BLACK * 0.771^TALL` | Lin 2023 Results, Covariate analysis and final PopPK model selection (printed equation) |
| V covariate equation | `445 * (BSA/1.2)^1.61` | Lin 2023 Results, Covariate analysis and final PopPK model selection (printed equation) |
| BSA reference 1.2 m^2 | n/a | Lin 2023 Results, Base model ("The BSA standard used for scaling was 1.2 m^2") |
| Mixed-order absorption structure | n/a | Lin 2023 Results, Base model; Discussion; Lin 2021 Results, Base model |
| One-compartment linear elimination | n/a | Lin 2023 Results, Base model |
| Proportional + additive residual error | n/a | Lin 2023 Table 2 and Results, Base model |

The `ini()` values reconcile with Table 2's own CV column, which is a useful
check that the Estimate column holds variances rather than standard deviations:

```{r iiv-reconcile}
sprintf("omega^2 = %.3f -> CV = %.1f%% (Table 2 prints %s)",
        c(0.125, 0.479),
        100 * sqrt(exp(c(0.125, 0.479)) - 1),
        c("36.5%", "78.4%"))
```

# Model structure

One compartment with linear elimination and **simultaneous mixed-order
absorption**: the dose enters the `depot` as a zero-order input running at the
constant rate `R1`, while first-order absorption `Ka` drains the depot into
`central` at the same time. Lin 2023 (Results, Base model) describes it as
"simultaneous mixed-order absorption (zero-order infusion followed by a
first-order absorption)"; the upstream Lin 2021 calls the same construct a
"sequential mixed order absorption function". `R1` is the NONMEM reserved
zero-order input rate for the dosing compartment, so the input duration is
`F * amt / R1`.

Because `Ka` (0.0369 /h) is roughly nine times smaller than
`kel = CL/V` (0.328 /h at the reference BSA), disposition is **absorption rate
limited (flip-flop)** and the terminal slope reports `Ka`, not `kel`.

## Dose records must set `rate = -1`

rxode2 only consults a modelled `rate()` when the dose record asks for it. With
the default `rate = 0` the modelled rate is **silently ignored** and the dose is
delivered as an instantaneous bolus into the depot. The check below documents
the difference so the requirement is visible rather than folklore.

```{r rate-semantics}
tv <- rxode2::zeroRe(mod)

solve_one <- function(dose, rate, bsa = 1.2, black = 0, tall = 0,
                      times = seq(0, 240, by = 0.05), ii = 0, addl = 0) {
  ev <- rxode2::et(amt = dose, cmt = "depot", rate = rate, ii = ii, addl = addl) |>
    rxode2::et(times, cmt = "central")
  ev <- as.data.frame(ev)
  ev$BSA <- bsa; ev$RACE_BLACK <- black; ev$DIS_TALL <- tall
  out <- rxode2::rxSolve(tv, ev, returnType = "data.frame")
  if (is.null(out$id)) out$id <- 1L
  out[!is.na(out$Cc), ]
}

probe_dose <- 20000                       # IU; see "Dose units" below
s_rate  <- solve_one(probe_dose, rate = -1)
s_bolus <- solve_one(probe_dose, rate = 0)

dur_expected <- exp(mod$theta[["lfdepot"]]) * probe_dose / exp(mod$theta[["lr1"]])
ka_i  <- exp(mod$theta[["lka"]])
kel_i <- 146 / 445
tmax_bolus_expected <- log(kel_i / ka_i) / (kel_i - ka_i)

rate_check <- data.frame(
  Quantity = c("depot peak time (h)", "Tmax of Cc (h)", "Cmax (IU/mL)"),
  `rate = -1` = c(s_rate$time[which.max(s_rate$depot)],
                  s_rate$time[which.max(s_rate$Cc)],
                  max(s_rate$Cc)),
  `rate = 0` = c(s_bolus$time[which.max(s_bolus$depot)],
                 s_bolus$time[which.max(s_bolus$Cc)],
                 max(s_bolus$Cc)),
  check.names = FALSE
)
knitr::kable(rate_check, digits = 3,
             caption = "Modelled zero-order rate is active only with rate = -1.")

sprintf("expected zero-order duration F*amt/R1 = %.3f h", dur_expected)
sprintf("expected bolus Tmax log(kel/ka)/(kel-ka) = %.3f h", tmax_bolus_expected)

# Deterministic identities, so these are asserted tightly.
stopifnot(
  abs(s_rate$time[which.max(s_rate$depot)] - dur_expected) < 0.06,
  abs(s_bolus$time[which.max(s_bolus$Cc)] - tmax_bolus_expected) < 0.06,
  s_bolus$time[which.max(s_bolus$depot)] == 0
)
```

Note that `Cmax` differs by well under 1% between the two, so **Cmax alone will
not reveal the mistake** -- only the depot peak time and Tmax do.

# Dose units: activity units, not mass

The model's amount unit is **IU of asparaginase activity, not mg**. Two
independent facts in the sources force this reading: the zero-order absorption
rate is reported in **IU/h** (Lin 2023 Table 2; Lin 2021 Table 2), and the
observation `Cc = central / vc` is only IU/mL if `central` is in IU.

**Neither Lin 2023 nor its upstream Lin 2021 states the mg-to-IU specific
activity of JZP458.** A clinical dose of 25 mg/m^2 therefore cannot be converted
into model units from any published source, and no conversion factor is baked
into the model file. Dose this model in IU.

The primary validation below is built entirely from **dose-independent**
published quantities, so it does not depend on resolving this gap. A
reconstruction of the dose scale is given afterwards, clearly separated and
flagged as not a published value.

# Validation 1: dose-independent post-hoc parameters

Lin 2023 (Results, Covariate analysis and final PopPK model selection) reports
three post-hoc quantities that do not depend on the dose scale: the geometric
mean apparent clearance `CL/F` of 0.31 L/h/m^2, apparent volume `V/F` of
1.14 L/m^2, and terminal half-life `t1/2` of 19.1 h. `CL/F` and `V/F` follow by
arithmetic from Table 2; `t1/2` is recovered from the simulated terminal slope.

```{r validation-postdoc}
bsa_ref <- 1.2
cl_ref  <- exp(mod$theta[["lcl"]])
vc_ref  <- exp(mod$theta[["lvc"]])
f_ref   <- exp(mod$theta[["lfdepot"]])

clf_per_m2 <- cl_ref / f_ref / bsa_ref / 1000   # mL/h -> L/h, then per m^2
vf_per_m2  <- vc_ref / f_ref / bsa_ref / 1000   # mL   -> L,   then per m^2

# Terminal slope taken well after the input has stopped, so it reports ka.
tail_s  <- s_rate[s_rate$time >= 150 & s_rate$time <= 240, ]
slope   <- stats::coef(stats::lm(log(tail_s$Cc) ~ tail_s$time))[[2]]
thalf   <- log(2) / (-slope)

posthoc <- data.frame(
  Quantity  = c("CL/F (L/h/m^2)", "V/F (L/m^2)", "Terminal t1/2 (h)"),
  Model     = c(clf_per_m2, vf_per_m2, thalf),
  Published = c(0.31, 1.14, 19.1)
)
posthoc$`Difference (%)` <- 100 * (posthoc$Model - posthoc$Published) / posthoc$Published
knitr::kable(posthoc, digits = 3,
             caption = "Model versus the post-hoc values reported by Lin 2023.")
```

The half-life is the sharpest of the three: it is a pure statement that the
terminal phase is absorption rate limited, and `log(2)/Ka = 18.78 h` reproduces
the published 19.1 h to within 2%. Lin 2023 reports only a 3% CV on the
individual `t1/2` values, which is itself evidence that `Ka` was fixed rather
than estimated. `CL/F` and `V/F` are geometric means over a population whose
BSA is normalised linearly in the paper's summary but enters the model as
`BSA^1.48` and `BSA^1.61`, so a few percent of disagreement is expected.

```{r validation-posthoc-assert}
# Deterministic (typical-value, no IIV), so these are tight.
stopifnot(
  abs(thalf - log(2) / ka_i) < 0.05,          # recovers log(2)/Ka exactly
  abs(thalf - 19.1) / 19.1 < 0.05,            # and matches the published value
  abs(clf_per_m2 - 0.31) / 0.31 < 0.15,
  abs(vf_per_m2 - 1.14) / 1.14 < 0.15
)
```

# Validation 2: exact covariate identities

The covariate model is multiplicative and acts only on clearance and volume, so
each published coefficient must reproduce exactly as a ratio of typical-value
exposures. These are deterministic, so they are asserted to solver tolerance.

```{r validation-covariates}
auc_of <- function(bsa, black, tall) {
  s <- solve_one(probe_dose, rate = -1, bsa = bsa, black = black, tall = tall,
                 times = seq(0, 600, by = 0.25))
  sum(diff(s$time) * (utils::head(s$Cc, -1) + utils::tail(s$Cc, -1)) / 2)
}
auc_ref <- auc_of(1.2, 0, 0)

cov_check <- data.frame(
  Comparison = c("Black/African American vs reference",
                 "T-ALL vs reference",
                 "BSA 2.4 vs 1.2 m^2 (dose held fixed)"),
  Observed = c(auc_of(1.2, 1, 0) / auc_ref,
               auc_of(1.2, 0, 1) / auc_ref,
               auc_of(2.4, 0, 0) / auc_ref),
  Expected = c(1 / 0.674, 1 / 0.771, 2^-1.48)
)
knitr::kable(cov_check, digits = 5,
             caption = "Exposure ratios reproduce the published coefficients.")

stopifnot(max(abs(cov_check$Observed - cov_check$Expected)) < 1e-4)
```

The AUC ratio is `1/0.674` and `1/0.771` rather than `0.674` and `0.771`
because these coefficients act on **clearance**, and AUC is inversely
proportional to clearance. The BSA row confirms the 1.48 exponent: holding the
dose fixed, doubling BSA divides exposure by `2^1.48`. Because the clinical dose
is itself BSA-proportional, the net effect of larger BSA on exposure is
`BSA^(1 - 1.48) = BSA^-0.48` -- a mild *decrease*, which is why the paper's
smallest patients are not the ones at risk of missing the NSAA target.

# Validation 3: PKNCA on the typical-value profile

NCA is run on the typical-value (no IIV, no residual error) single-dose profile.
Running it over a full-IIV cohort would make the half-life estimate both
NA-poisoned and Tmax-selected, which is not what the published `t1/2` means.

```{r pknca}
nca_in <- s_rate |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc)

# Defensive time-zero record so PKNCA does not warn about an AUC range
# starting before the first measurement.
if (!any(nca_in$time == 0)) {
  nca_in <- dplyr::bind_rows(
    data.frame(id = 1L, time = 0, Cc = 0), nca_in
  ) |> dplyr::distinct(id, time, .keep_all = TRUE) |> dplyr::arrange(time)
}

dose_df <- data.frame(id = 1L, time = 0, amt = probe_dose)

o_conc <- PKNCA::PKNCAconc(nca_in, Cc ~ time | id)
o_dose <- PKNCA::PKNCAdose(dose_df, amt ~ time | id)
o_data <- PKNCA::PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(
    start = 0, end = 240,
    cmax = TRUE, tmax = TRUE, half.life = TRUE,
    auclast = TRUE, aucinf.obs = TRUE
  )
)
res <- as.data.frame(PKNCA::pk.nca(o_data))
knitr::kable(res[, c("PPTESTCD", "PPORRES")], digits = 4,
             caption = "PKNCA summary of the typical-value single-dose profile.")
```

```{r pknca-compare}
get_nca <- function(code) res$PPORRES[res$PPTESTCD == code]
stopifnot(length(get_nca("half.life")) == 1L)   # guard: the lookup must hit

simulated <- data.frame(
  `t.5` = get_nca("half.life"),
  Tmax  = get_nca("tmax"),
  check.names = FALSE
)
reference <- data.frame(
  `t.5` = 19.1,
  Tmax  = 12.0,
  check.names = FALSE
)
knitr::kable(
  nlmixr2lib::ncaComparisonTable(
    simulated, reference,
    units = c(`t.5` = "h", Tmax = "h")
  ),
  caption = paste(
    "Typical-value NCA versus the median post-hoc values of Lin 2023.",
    "Tmax depends on the dose scale; see Errata."
  )
)
```

`aucinf.obs` equals `auclast` here because the 240 h window is roughly 13
absorption half-lives, so the extrapolated tail is negligible; the identity is
asserted as an inequality rather than an equality.

```{r pknca-identities}
stopifnot(
  get_nca("aucinf.obs") >= get_nca("auclast"),
  abs(get_nca("half.life") - log(2) / ka_i) < 0.5,
  abs(get_nca("cmax") - max(s_rate$Cc)) < 1e-6
)
```

# Dose-scale reconstruction (not a published value)

Everything above is dose-independent. To simulate the clinical mg/m^2 regimens
the missing mg-to-IU factor is needed, so this section **reconstructs** it from
the paper's own simulation output and labels the result clearly. It is not a
published number and is not part of the model file.

Lin 2023 Figure 3b plots the simulated median SAA profile for
25 mg/m^2 every 48 h in the NHANES virtual population (median BSA 1.66 m^2).
Reading the **first peak** off that figure at about 1.0 IU/mL fixes the scale;
the steady-state peak, the trough and Tmax are then **held-out** landmarks that
the reconstruction was not fitted to.

```{r dose-scale}
bsa_nhanes <- 1.66
mgm2       <- 25

first_peak_target <- 1.0   # digitised from Lin 2023 Figure 3b

sim_regimen <- function(dose_iu, bsa) {
  solve_one(dose_iu, rate = -1, bsa = bsa,
            times = seq(0, 336, by = 0.25), ii = 48, addl = 6)
}
obj <- function(dose_iu) {
  s <- sim_regimen(dose_iu, bsa_nhanes)
  max(s$Cc[s$time <= 48]) - first_peak_target
}
dose_iu_recon <- stats::uniroot(obj, c(2000, 200000))$root
specific_activity <- dose_iu_recon / (mgm2 * bsa_nhanes)

sprintf("reconstructed specific activity ~ %.0f IU/mg (NOT a published value)",
        specific_activity)

s_rec <- sim_regimen(dose_iu_recon, bsa_nhanes)
landmarks <- data.frame(
  Landmark = c("first peak (fitted)", "steady-state peak (held out)",
               "trough before dose 7 (held out)", "Tmax of first dose (held out)"),
  Model = c(max(s_rec$Cc[s_rec$time <= 48]),
            max(s_rec$Cc[s_rec$time > 240 & s_rec$time < 290]),
            s_rec$Cc[which.min(abs(s_rec$time - 288))],
            s_rec$time[which.max(s_rec$Cc[s_rec$time <= 48])]),
  `Lin 2023` = c("~1.0 IU/mL (Figure 3b)", "~1.3 IU/mL (Figure 3b)",
                 "~0.35-0.40 IU/mL (Figure 3b)", "12.0 h median (Results text)"),
  check.names = FALSE
)
knitr::kable(landmarks, digits = 2,
             caption = "Held-out landmarks the dose-scale fit was not calibrated to.")
```

```{r dose-scale-plot, fig.width = 7, fig.height = 4}
ggplot2::ggplot(s_rec, ggplot2::aes(time, Cc)) +
  ggplot2::geom_line(colour = "steelblue") +
  ggplot2::geom_hline(yintercept = 0.1, linetype = "dashed") +
  ggplot2::scale_y_log10() +
  ggplot2::scale_x_continuous(breaks = seq(0, 336, by = 48)) +
  ggplot2::labs(
    x = "Time (hours)", y = "SAA (IU/mL)",
    title = "Typical-value SAA, 25 mg/m^2 every 48 h at BSA 1.66 m^2",
    subtitle = "Compare Lin 2023 Figure 3b (median of the NHANES virtual population)"
  ) +
  ggplot2::theme_bw()
```

```{r dose-scale-assert}
# Loose bounds: the target was read off a log-scale figure by eye, so the
# reconstruction carries perhaps 20-25% uncertainty. These bounds still go red
# on a mis-transcribed CL, V, Ka or R1, which move the landmarks severalfold.
stopifnot(
  specific_activity > 350, specific_activity < 900,
  max(s_rec$Cc[s_rec$time > 240 & s_rec$time < 290]) > 1.0,
  max(s_rec$Cc[s_rec$time > 240 & s_rec$time < 290]) < 1.7,
  s_rec$Cc[which.min(abs(s_rec$time - 288))] > 0.2,
  s_rec$Cc[which.min(abs(s_rec$time - 288))] < 0.6
)
```

The reconstruction lands near 580 IU/mg. It rests on a single value read by eye
off a base-10 log axis on which 1.00 is a printed gridline, so it is worth
perhaps one significant figure. **No conversion factor is committed to the
model file**: the specific activity of JZP458 is a product attribute rather than
a model parameter, it is absent from every on-disk source, and baking a
figure-derived estimate into `ini()` would give it a false precision. See
Errata.

# Virtual cohort

A 200-subject cohort matching the AALL1931 analysis-set covariate distributions
is simulated on the 25 mg/m^2 every-48-h regimen, using the reconstructed dose
scale. This is illustrative rather than a reproduction of the paper's response
rates, which were computed on the NHANES virtual population.

```{r cohort}
rxode2::rxSetSeed(20230501)
n_sub <- 200

set.seed(20230501)
cohort <- data.frame(
  id  = seq_len(n_sub),
  # Log-normal BSA matched to the analysis-set median 1.17 and range 0.44-2.53.
  BSA = pmin(pmax(stats::rlnorm(n_sub, log(1.17), 0.42), 0.44), 2.53),
  RACE_BLACK = stats::rbinom(n_sub, 1, 0.133),
  DIS_TALL   = stats::rbinom(n_sub, 1, 0.157)
)
cohort$amt <- mgm2 * cohort$BSA * specific_activity

ev_cohort <- lapply(seq_len(n_sub), function(i) {
  e <- rxode2::et(amt = cohort$amt[i], cmt = "depot", rate = -1,
                  ii = 48, addl = 6) |>
    rxode2::et(seq(0, 336, by = 2), cmt = "central")
  e <- as.data.frame(e)
  e$id <- cohort$id[i]
  e$BSA <- cohort$BSA[i]
  e$RACE_BLACK <- cohort$RACE_BLACK[i]
  e$DIS_TALL <- cohort$DIS_TALL[i]
  e
}) |> dplyr::bind_rows()

sim_cohort <- rxode2::rxSolve(mod, ev_cohort, returnType = "data.frame")
if (is.null(sim_cohort$id)) sim_cohort$id <- 1L
sim_cohort <- sim_cohort[!is.na(sim_cohort$Cc), ]

# Nadir SAA over the final 48 h dosing interval (after the 7th dose).
nsaa <- sim_cohort |>
  dplyr::filter(time >= 288, time <= 336) |>
  dplyr::group_by(id) |>
  dplyr::summarise(nsaa = min(Cc), .groups = "drop")

sprintf("median NSAA over the last interval = %.2f IU/mL", stats::median(nsaa$nsaa))
sprintf("proportion with NSAA >= 0.1 IU/mL = %.1f%% (Lin 2023 reports 93.8%% on the NHANES population)",
        100 * mean(nsaa$nsaa >= 0.1))
```

This cohort clears the 0.1 IU/mL target more often than the 93.8% Lin 2023
reports, for two structural reasons rather than a disagreement with the model.
First, exposure scales as `BSA^(1 - 1.48) = BSA^-0.48`, and the AALL1931
analysis set is much smaller-bodied (median BSA 1.17 m^2) than the general
population NHANES cohort the paper simulated (median 1.66 m^2, spanning ages
1 month to 80 years), so the same mg/m^2 dose yields higher activity here.
Second, `Cc` is the individual prediction and carries no residual error, whereas
a simulated *observation* of a nadir near the assay's lower limit is pushed
below the threshold by the proportional error (SD 0.46) a meaningful fraction of
the time. The number below is therefore an optimistic bound on the response
rate, not a reproduction of it.

```{r cohort-plot, fig.width = 7, fig.height = 4}
ribbon <- sim_cohort |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    lo  = stats::quantile(Cc, 0.025),
    med = stats::median(Cc),
    hi  = stats::quantile(Cc, 0.975),
    .groups = "drop"
  )

ggplot2::ggplot(ribbon, ggplot2::aes(time)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi), alpha = 0.2,
                       fill = "steelblue") +
  ggplot2::geom_line(ggplot2::aes(y = med), colour = "steelblue") +
  ggplot2::geom_hline(yintercept = 0.1, linetype = "dashed") +
  ggplot2::scale_y_log10() +
  ggplot2::scale_x_continuous(breaks = seq(0, 336, by = 48)) +
  ggplot2::labs(
    x = "Time (hours)", y = "SAA (IU/mL)",
    title = "Simulated SAA, 25 mg/m^2 every 48 h, 200 virtual AALL1931 patients",
    subtitle = "Median with 95% prediction interval; dashed line is the 0.1 IU/mL NSAA target"
  ) +
  ggplot2::theme_bw()
```

```{r cohort-assert}
# Cohort-derived, so assert on the CENTRE and a robust quantile only -- never on
# the extremes of a random cohort, which are not reproducible across rxode2
# builds or solver thread counts.
stopifnot(
  stats::median(nsaa$nsaa) > 0.15,
  stats::median(nsaa$nsaa) < 1.0,
  mean(nsaa$nsaa >= 0.1) > 0.6
)
```

# Assumptions and deviations / Errata

1. **The model's amount unit is IU, and no mg-to-IU conversion is published.**
   The zero-order absorption rate is reported in IU/h in both Lin 2023 Table 2
   and Lin 2021 Table 2, and `Cc = central / vc` is IU/mL only if the dose is in
   IU. Neither publication states the specific activity of JZP458, so a
   25 mg/m^2 dose cannot be expressed in model units from any published source.
   **Dose this model in IU.** No conversion factor is present in the model file.

2. **The dose scale used below the "Dose-scale reconstruction" heading is
   figure-derived, not published.** It is obtained by calibrating the model to
   one landmark digitised from Lin 2023 Figure 3b (the first peak of the
   25 mg/m^2 every-48-h median profile, about 1.0 IU/mL), which implies roughly
   580 IU/mg. That calibration then reproduces three landmarks it was *not*
   fitted to -- the steady-state peak and trough of the same figure, and the
   order of magnitude of the Results-text Tmax -- which is the only evidence
   offered for it. Because it is a single eyeball read off a log axis, it is
   good to about one significant figure and is used only for illustration; it
   is flagged inline wherever it is used and appears nowhere in the model file.

3. **Absorption structure was inferred from prose plus the parameter set, not
   from a printed equation.** Lin 2023 prints the two covariate equations but no
   absorption equation. The encoding here -- a zero-order input filling the
   depot at `R1` while first-order `Ka` drains it -- is the reading that (a)
   matches "zero-order infusion followed by a first-order absorption", (b) uses
   exactly the three reported fixed parameters `Ka`, `R1` and `F`, (c) matches
   NONMEM's reserved `R1` rate-for-compartment-1 semantics, and (d) is the only
   reading that reproduces the published terminal half-life. A parallel
   zero-plus-first-order absorption *out of* the depot was rejected because the
   depot then empties in finite time and the terminal half-life collapses to
   `log(2)/kel = 2.1 h` instead of the reported 19.1 h.

4. **Median Tmax is only approximately reproduced, and only under the
   reconstructed dose scale.** Lin 2023 reports a median individual predicted
   Tmax of 12.0 h with a range of 4-24 h. The value the model gives at the
   reconstructed scale is printed in the held-out-landmarks table above and sits
   inside that range. Tmax scales with the dose through the zero-order input
   duration `F * amt / R1`, so it is not a dose-independent check and it cannot
   be used to arbitrate the scale. Note also that every reported Tmax value
   (median 12.0, range 4 and 24) is a multiple of 4 h, which suggests the
   post-hoc Tmax was read off a 4-hourly output grid rather than a continuous
   optimum.

5. **Table 2 labels the bioavailability row "(%)" but the value is a
   fraction.** `F = 0.359` is 35.9%, not 0.359%. Two checks in the upstream
   Lin 2021 settle it: its Table 2 gives `F = 0.365` while its Results text
   calls the same quantity "bioavailability at 36.5%", and its Table 2 footnote
   reconciles `Vd = 3030 mL` (i.v.) against `Vd/F = 8.30 L` (i.m.), which
   requires `3030 / 0.365 = 8301 mL`. Taking the header literally would inflate
   the apparent volume a hundredfold, so the fraction reading is encoded.

6. **`$SIGMA` fixed to 1 is not a third error term.** Lin 2023 Table 2's final
   row, "Within-subject variability | Proportional | 1 | 100%", is the NONMEM
   idiom of fixing `$SIGMA` to 1 so the residual magnitudes are carried by the
   two `$THETA` rows above it. It is encoded here as `propSd = 0.46` and
   `addSd = 0.0166` with no additional parameter.

7. **No CL-V correlation is encoded.** Lin 2023 reports IIV on clearance and
   volume separately and no covariance, so the two etas are left uncorrelated.

8. **T-ALL is not T-lineage.** The retained covariate is the T-cell *ALL*
   subtype only. The 9.6% of patients with T-cell lymphoblastic *lymphoma* are
   in the reference group, because Lin 2023 screened primary disease (ALL vs
   LBL) and disease subtype (B vs T cell) as separate covariates and retained
   only the T-ALL cell of that cross-classification.

9. **Table 1 misprints the age range.** Lin 2023 Table 1 gives age as
   "10.0 (1.7-2.5)", which contradicts both the same table's mean of 10.2 years
   and the Results text, "ranged from 1.7 to 25 years of age". The model's
   `population` metadata records the text value, 1.7-25 years.

10. **Screened-but-not-retained covariates are documented, not modelled.** Age,
   sex, height, body weight, ethnicity, primary disease and ADA status were all
   evaluated and rejected. They are recorded in the model file's
   `covariatesDataExcluded` list so the composition of the covariate screen is
   auditable without introducing unused `covariateData` entries.

11. **The virtual cohort is illustrative.** Lin 2023's published response rates
    were computed on a 2000-subject NHANES virtual population constructed to
    match real-world ALL race epidemiology. The 200-subject cohort here matches
    the AALL1931 analysis-set covariate distributions instead, and additionally
    depends on the reconstructed dose scale, so its response rate is not a
    like-for-like reproduction of the published 93.8%.
