---
title: "Glasdegib (Lin 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Glasdegib (Lin 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Models and source

Lin 2020 developed a population PK model for oral glasdegib in patients with
cancer. Table 3 of the paper also reports, for comparison, a separately fitted
model from healthy volunteers. The two models were fitted to different
cohorts, so the paper contributes two model files.

```{r models}
mod_pat <- rxode2::rxode(readModelDb("Lin_2020_glasdegib"))
mod_hv <- rxode2::rxode(readModelDb("Lin_2020_glasdegib_healthy"))

tibble::tibble(
  Model = c("Lin_2020_glasdegib", "Lin_2020_glasdegib_healthy"),
  Cohort = c(
    "269 patients with hematologic malignancies or solid tumors (final model)",
    "49 healthy volunteers (exploratory comparison model, Table 3 footnote b)"
  )
) |>
  knitr::kable()
```

* Citation: `r mod_pat$reference`
* Article: <https://doi.org/10.1002/jcph.1556> (open access, PMC7187372)
* Supplement (Supplemental Tables S1-S5, Figures S1-S2):
  <https://www.ebi.ac.uk/europepmc/webservices/rest/PMC7187372/supplementaryFiles>

Both models are two-compartment models with first-order absorption. The
patient model's covariates are baseline body weight (allometric), the
weight-standardized creatinine clearance, the baseline percentage of bone
marrow blasts and concomitant CYP3A inhibitors on CL/F, and tumor type (solid
vs hematologic) on Vp/F and Q/F.

The steady-state exposures from this model were used in the exposure-response
analyses of the same group; the overall-survival models from that work are
packaged as `Lin_2020_glasdegib_treatment`, `Lin_2020_glasdegib_exposure` and
`Lin_2020_glasdegib_decitabine`.

# Population

The analysis pooled three studies: B1371001 (47 patients with hematologic
malignancies, 5-600 mg QD), B1371002 (23 patients with solid tumors, 80-640 mg
QD) and B1371003 (202 patients with AML or high-risk MDS, 100 or 200 mg QD
combined with low-dose cytarabine, decitabine or 7 + 3 chemotherapy). Of the
272 enrolled patients, 3 had no PK samples, which left 269 patients (246
hematologic, 23 solid tumor) and 3616 concentrations. Table 2 gives the
baseline characteristics of all 272: median age 69 years (25-92), median
weight 78.6 kg (43.5-145.6), 181 male and 91 female, 86% White. Median
Cockcroft-Gault creatinine clearance was 80.9 mL/min (31.4-238.4), and 62 and
56 patients took a moderate or a strong CYP3A inhibitor. 187 patients (69%)
received the clinical dose of 100 mg QD.

The healthy-volunteer model used 1937 concentrations from 49 noncancer
subjects of studies B1371010 (a ketoconazole interaction study with a 200 mg
single dose of glasdegib) and B1371014. The paper does not report their
demographics.

```{r population}
str(readModelDb("Lin_2020_glasdegib")()$population[c("n_subjects", "age_range", "weight_range")])
```

# Source trace

Every `ini()` value has a comment in the model file giving its source. The
covariate coefficients are taken from the final-model equations in Results
("Covariate Analyses and Final Model"), which print more digits than Table 3.

| Parameter | Value | Source |
|---|---|---|
| `lka` | log(0.06) | Table 3, ka = 0.06 1/h |
| `lcl` | log(6.27) | Table 3 and CL/F equation |
| `lvc` | log(3.32) | Table 3 and Vc/F equation |
| `lvp` | log(279.21) | Table 3 and Vp/F equation |
| `lq` | log(1.29) | Table 3 and Q/F equation |
| `e_wt_cl_q`, `e_wt_vc_vp` | 0.75, 1 (fixed) | Results "Base Model"; final-model equations |
| `e_crcl_cl` | 0.406 | CL/F equation, `(WNCL/71.23)^0.406` |
| `e_bmblast_pct_cl` | -0.004 | CL/F equation, `(1 - 0.004 (BPBL - 38.20))` |
| `e_cyp3a4_inh_mod_cl` | -0.173 | CL/F equation, `(1 - 0.173 CYPmoderate)` |
| `e_cyp3a4_inh_strong_cl` | -0.303 | CL/F equation, `(1 - 0.303 CYPstrong)` |
| `e_tumtp_solid_vp` | -0.825 | Vp/F equation, `(1 - 0.825 Solid)` |
| `e_tumtp_solid_q` | -0.653 | Q/F equation, `(1 - 0.653 Solid)` |
| IIV CL/F, Vc/F, Vp/F, Q/F, ka | 43.0, 215.3, 112.8, 65.8, 13.5 %CV | Table 3; `omega^2 = log(1 + CV^2)` |
| `expSdHeme`, `expSdSolid` | 0.658, 0.595 | Table 3 residual error; Methods (log-transformed data, thetarized sigma) |
| WNCL = CRCL x 70 / WT | n/a | Table 1 footnote a |
| Two-compartment ODEs, first-order absorption | n/a | Methods; Figure 1 |
| Healthy-volunteer model: CL/F, Vc/F, Vp/F, Q/F, ka | 10.1, 112, 53.5, 1.92, 0.79 | Table 3, "Healthy Volunteer Model" column |
| Healthy-volunteer IIV CL/F, Vc/F, Vp/F, Q/F, ka | 19.6, 11.0, 3.2 (fixed), 3.2 (fixed), 53.4 %CV | Table 3 |
| Healthy-volunteer proportional residual error | 58.6% | Table 3, footnote b |

# Typical-value covariate effects (Table 4)

Table 4 lists the typical CL/F, Vc/F, Vp/F and Q/F for a 70 kg hematologic
patient with 38.2% marrow blasts and no CYP3A inhibitor, then changes one
covariate at a time. The chunk below computes the same quantities from the
packaged model: it solves the model with the random effects set to zero and
reads the individual parameters from the output.

```{r table4}
mod_pat_typ <- rxode2::zeroRe(mod_pat)

# Reference: WT 70 kg and WNCL = 71.23 mL/min, i.e. raw CRCL = 71.23 at 70 kg.
t4 <- tibble::tribble(
  ~scenario, ~WT, ~wncl, ~BMBLAST_PCT, ~CONMED_CYP3A4_INH_MOD, ~CONMED_CYP3A4_INH_STRONG, ~TUMTP_SOLID,
  "Typical patient", 70, 71.23, 38.2, 0, 0, 0,
  "Body weight 61.24 kg", 61.24, 71.23, 38.2, 0, 0, 0,
  "Body weight 102.1 kg", 102.1, 71.23, 38.2, 0, 0, 0,
  "Marrow blasts 15%", 70, 71.23, 15, 0, 0, 0,
  "Marrow blasts 83%", 70, 71.23, 83, 0, 0, 0,
  "Moderate CYP3A inhibitor", 70, 71.23, 38.2, 1, 0, 0,
  "Strong CYP3A inhibitor", 70, 71.23, 38.2, 0, 1, 0,
  "Solid tumor", 70, 71.23, 38.2, 0, 0, 1
) |>
  mutate(id = row_number(), CRCL = wncl * WT / 70)

ev_t4 <- t4 |>
  select(id, WT, CRCL, BMBLAST_PCT, CONMED_CYP3A4_INH_MOD, CONMED_CYP3A4_INH_STRONG, TUMTP_SOLID) |>
  mutate(time = 0, evid = 0, amt = 0, cmt = "central")

par_t4 <- rxode2::rxSolve(mod_pat_typ, events = ev_t4, returnType = "data.frame") |>
  select(id, cl, vc, vp, q) |>
  left_join(t4 |> select(id, scenario), by = "id")

# Published Table 4 values (rounded to 0.1 in the paper); NA where Table 4 has
# no entry for that parameter in that scenario.
pub_t4 <- tibble::tribble(
  ~scenario, ~cl_pub, ~vc_pub, ~vp_pub, ~q_pub,
  "Typical patient", 6.3, 3.3, 279.2, 1.3,
  "Body weight 61.24 kg", 5.7, 2.9, NA, NA,
  "Body weight 102.1 kg", 8.3, 4.8, NA, NA,
  "Marrow blasts 15%", 6.9, NA, NA, NA,
  "Marrow blasts 83%", 5.1, NA, NA, NA,
  "Moderate CYP3A inhibitor", 5.2, NA, NA, NA,
  "Strong CYP3A inhibitor", 4.4, NA, NA, NA,
  "Solid tumor", NA, NA, 48.9, 0.4
)

t4_cmp <- par_t4 |> left_join(pub_t4, by = "scenario")

t4_cmp |>
  select(scenario, cl, cl_pub, vc, vc_pub, vp, vp_pub, q, q_pub) |>
  dplyr::rename(
    Scenario = scenario,
    "CL/F model" = cl, "CL/F Table 4" = cl_pub,
    "Vc/F model" = vc, "Vc/F Table 4" = vc_pub,
    "Vp/F model" = vp, "Vp/F Table 4" = vp_pub,
    "Q/F model" = q, "Q/F Table 4" = q_pub
  ) |>
  knitr::kable(digits = 2, caption = "Replicates Table 4 of Lin 2020 (L/h and L).")

# Deterministic check: the model values must round to the published values.
# Table 4 prints one decimal place, so the tolerance is half a unit in that
# place (0.05) plus a little slack for the paper's own rounding of inputs.
long_cmp <- bind_rows(
  t4_cmp |> transmute(scenario, model = cl, pub = cl_pub),
  t4_cmp |> transmute(scenario, model = vc, pub = vc_pub),
  t4_cmp |> transmute(scenario, model = vp, pub = vp_pub),
  t4_cmp |> transmute(scenario, model = q, pub = q_pub)
) |>
  filter(!is.na(pub))
stopifnot(nrow(long_cmp) == 14, all(abs(long_cmp$model - long_cmp$pub) <= 0.06))
```

All fourteen Table 4 entries are reproduced to the printed precision. Table 4
also lists CL/F by renal-function group (6.5, 6.2 and 4.8 L/h at median raw
CRCL 110.2, 75.5 and 51.1 mL/min). Its footnote and the Discussion describe
these as body-weight-normalized post hoc estimates, not typical values from
the equation, so they are not a check on the equation and are left out of the
comparison above.

# Steady-state exposure by covariate scenario (Figure 4)

Figure 4 shows the median steady-state Cmax and AUCtau for each covariate
scenario (500 simulated patients each). The caption does not state the dose;
the clinical dose of 100 mg QD is assumed, and the agreement below supports
it. The caption describes the reference as "a 70-kg male hematology patient"
but also as having "the median values for baseline weight". The printed
numbers fit the median, 78.6 kg: at 70 kg every non-weight row would be about
9% above the printed value. Sex is not a covariate, so the "Female" row is the
reference patient: 78.6 kg, median WNCL and median marrow blasts, hematologic
malignancy and no CYP3A inhibitor. The typical-value model below uses these
reference covariates and changes one of them per row. PKNCA computes Cmax and
AUC over the dosing interval on day 60.

```{r fig4-sim}
fig4 <- tibble::tribble(
  ~scenario, ~WT, ~wncl, ~BMBLAST_PCT, ~CONMED_CYP3A4_INH_MOD, ~CONMED_CYP3A4_INH_STRONG, ~TUMTP_SOLID, ~cmax_pub, ~auc_pub,
  "Female (reference)", 78.6, 71.23, 38.2, 0, 0, 0, 995.5, 14980.2,
  "Solid tumor", 78.6, 71.23, 38.2, 0, 0, 1, 1021.4, 15269.9,
  "Strong CYP inhibitor", 78.6, 71.23, 38.2, 0, 1, 0, 1327.1, 20317.5,
  "Moderate CYP inhibitor", 78.6, 71.23, 38.2, 1, 0, 0, 1103.0, 16829.0,
  "BPBL 80.0", 78.6, 71.23, 80.0, 0, 0, 0, 1099.1, 16759.5,
  "BPBL 60.0", 78.6, 71.23, 60.0, 0, 0, 0, 1043.3, 15749.1,
  "BPBL 25.0", 78.6, 71.23, 25.0, 0, 0, 0, 922.4, 13988.1,
  "BPBL 17.8", 78.6, 71.23, 17.8, 0, 0, 0, 904.9, 13512.3,
  "WNCL 109.59", 78.6, 109.59, 38.2, 0, 0, 0, 821.2, 12250.7,
  "WNCL 88.48", 78.6, 88.48, 38.2, 0, 0, 0, 899.6, 13538.4,
  "WNCL 57.45", 78.6, 57.45, 38.2, 0, 0, 0, 1047.4, 15800.2,
  "WNCL 46.95", 78.6, 46.95, 38.2, 0, 0, 0, 1126.3, 17102.6,
  "BWT 102.28", 102.28, 71.23, 38.2, 0, 0, 0, 748.0, 11371.7,
  "BWT 89.00", 89.00, 71.23, 38.2, 0, 0, 0, 853.2, 12874.9,
  "BWT 78.60", 78.60, 71.23, 38.2, 0, 0, 0, 871.8, 13285.4,
  "BWT 68.00", 68.00, 71.23, 38.2, 0, 0, 0, 1017.1, 15327.3,
  "BWT 61.26", 61.26, 71.23, 38.2, 0, 0, 0, 1099.0, 16480.8
) |>
  mutate(id = row_number(), CRCL = wncl * WT / 70)

tau <- 24
ss_start <- 59 * tau
obs_times <- ss_start + seq(0, tau, by = 0.25)

ev_fig4 <- bind_rows(
  fig4 |> select(id) |>
    tidyr::crossing(time = seq(0, ss_start, by = tau)) |>
    mutate(evid = 1, amt = 100, cmt = "depot"),
  fig4 |> select(id) |>
    tidyr::crossing(time = obs_times) |>
    mutate(evid = 0, amt = 0, cmt = "central")
) |>
  left_join(fig4 |> select(id, scenario, WT, CRCL, BMBLAST_PCT, CONMED_CYP3A4_INH_MOD,
                           CONMED_CYP3A4_INH_STRONG, TUMTP_SOLID), by = "id") |>
  arrange(id, time, desc(evid))

sim_fig4 <- rxode2::rxSolve(mod_pat_typ, events = ev_fig4, keep = "scenario",
                            returnType = "data.frame")
```

```{r fig4-nca}
conc4 <- sim_fig4 |>
  dplyr::filter(!is.na(Cc)) |>
  select(id, time, Cc, scenario)
dose4 <- ev_fig4 |>
  dplyr::filter(evid == 1) |>
  select(id, time, amt, scenario)

nca4 <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc4, Cc ~ time | scenario + id),
  PKNCA::PKNCAdose(dose4, amt ~ time | scenario + id),
  intervals = data.frame(start = ss_start, end = ss_start + tau, cmax = TRUE, auclast = TRUE)
))

res4 <- as.data.frame(nca4) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "auclast")) |>
  select(scenario, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  left_join(fig4 |> select(scenario, cmax_pub, auc_pub), by = "scenario") |>
  mutate(
    cmax_pct = 100 * (cmax / cmax_pub - 1),
    auc_pct = 100 * (auclast / auc_pub - 1)
  )

res4 |>
  select(scenario, cmax, cmax_pub, cmax_pct, auclast, auc_pub, auc_pct) |>
  dplyr::rename(
    Scenario = scenario,
    "Cmax,ss model (ng/mL)" = cmax, "Cmax,ss Figure 4" = cmax_pub, "Cmax diff (%)" = cmax_pct,
    "AUCtau,ss model (ng*h/mL)" = auclast, "AUCtau,ss Figure 4" = auc_pub, "AUC diff (%)" = auc_pct
  ) |>
  knitr::kable(digits = 1, caption = "Replicates the printed medians of Figure 4 of Lin 2020 (100 mg QD, day 60).")
```

The CYP3A, marrow-blast, renal-function and tumor-type rows agree with the
printed medians to within about 5%. The body-weight rows are 3-11% higher
than printed. Figure 4 has its own internal noise of about that size. The
"Female" row and the "BWT 78.60" row describe the same patient, since sex is
not a covariate and 78.6 kg is the median weight, yet the figure prints AUCs
of 14980 and 13285 ng*h/mL for them, 11% apart. Each scenario appears to be a
separate 500-patient Monte Carlo simulation. The median AUC over log-normal
clearance is the typical value, so the typical-value solve here has no such
noise. The fold changes relative to the "Female" row are:

```{r fig4-fold}
ref_row <- res4 |> dplyr::filter(scenario == "Female (reference)")
fold4 <- res4 |>
  mutate(
    fold_model = auclast / ref_row$auclast,
    fold_pub = auc_pub / ref_row$auc_pub,
    fold_pct = 100 * (fold_model / fold_pub - 1)
  )
fold4 |>
  select(scenario, fold_model, fold_pub, fold_pct) |>
  dplyr::rename(
    Scenario = scenario, "AUC fold, model" = fold_model,
    "AUC fold, Figure 4" = fold_pub, "Difference (%)" = fold_pct
  ) |>
  knitr::kable(digits = 3, caption = "AUCtau,ss fold change relative to the reference patient.")

# Typical-value solve, so the model side is deterministic (no rxode2 RNG);
# the tolerance covers the Monte Carlo noise in the published medians, which
# the Female / BWT 78.60 pair puts at about 11%. Realised on authoring: median
# AUC difference +1.6%, median Cmax difference +0.8%, 90th percentile of the
# absolute fold difference about 9.2%. A mis-transcribed clearance, dose or
# unit moves these by tens of percent.
stopifnot(
  abs(median(res4$auc_pct)) < 10,
  abs(median(res4$cmax_pct)) < 10,
  quantile(abs(fold4$fold_pct), 0.9) < 15
)
```

The fold changes all point in the published direction and have the published
magnitudes. The largest differences are in the body-weight rows, which inherit
the offset described above. The CYP3A rows follow the equation exactly:
1 / (1 - 0.303) = 1.43-fold and 1 / (1 - 0.173) = 1.21-fold. Figure 4 prints
1.36 and 1.12. The Results and Table 4 quote the same 17% and 30% CL/F
reductions as the equation, so the equation is carried as published.

# Concentration-time profiles (Figure 3)

Figure 3 is a prediction-corrected VPC of the observed data by study, which
cannot be reproduced without the data. The chunk below shows the simulated
5th, 50th and 95th percentiles for 100 mg QD, the clinical dose, over the
first day and at steady state, for 200 hematologic and 200 solid-tumor
patients.

```{r vpc-cohort}
rxode2::rxSetSeed(20200501)
set.seed(20200501)

make_cohort <- function(n, solid, id_offset) {
  tibble(
    id = id_offset + seq_len(n),
    WT = pmin(pmax(exp(rnorm(n, log(78.6), 0.22)), 43.5), 145.6),
    CRCL = pmin(pmax(exp(rnorm(n, log(80.9), 0.35)), 31.4), 238.4),
    BMBLAST_PCT = if (solid) 38.2 else pmin(pmax(rnorm(n, 39.3, 25), 0), 100),
    CONMED_CYP3A4_INH_MOD = 0,
    CONMED_CYP3A4_INH_STRONG = 0,
    TUMTP_SOLID = as.integer(solid),
    group = if (solid) "Solid tumor" else "Hematologic"
  )
}
cohort <- bind_rows(
  make_cohort(200, solid = FALSE, id_offset = 0L),
  make_cohort(200, solid = TRUE, id_offset = 200L)
)
stopifnot(!anyDuplicated(cohort$id))

vpc_times <- c(seq(0, 24, by = 1), ss_start + seq(0, 24, by = 1))
ev_vpc <- bind_rows(
  cohort |> tidyr::crossing(time = seq(0, ss_start, by = tau)) |>
    mutate(evid = 1, amt = 100, cmt = "depot"),
  cohort |> tidyr::crossing(time = vpc_times) |>
    mutate(evid = 0, amt = 0, cmt = "central")
) |>
  arrange(id, time, desc(evid))

sim_vpc <- rxode2::rxSolve(mod_pat, events = ev_vpc, keep = "group", returnType = "data.frame")
```

```{r vpc-plot}
sim_vpc |>
  mutate(day = ifelse(time < ss_start, "Day 1", "Day 60 (steady state)"),
         tad = ifelse(time < ss_start, time, time - ss_start)) |>
  group_by(group, day, tad) |>
  summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::filter(tad > 0) |>
  ggplot(aes(tad, Q50, colour = group, fill = group)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line() +
  facet_wrap(~day) +
  scale_y_log10() +
  labs(x = "Time after dose (h)", y = "Glasdegib Cc (ng/mL)", colour = NULL, fill = NULL,
       caption = "Simulated 5th-95th percentiles, 100 mg QD; compare Figure 3 of Lin 2020.")
```

Figure 3C, study B1371003 where most patients took 100 mg QD, has its
prediction-corrected median at roughly 500-1000 ng/mL, with most observations
between about 100 and 3000 ng/mL. The simulated profiles fall in the same
range.

# Healthy-volunteer model

The Discussion states that the healthy-volunteer CL/F is "over 1.6-fold
higher" and ka "over 13-fold higher" than in patients. Both follow from the
two parameter sets:

```{r hv-ratios}
hv_ratio <- c(
  cl = exp(mod_hv$theta[["lcl"]] - mod_pat$theta[["lcl"]]),
  ka = exp(mod_hv$theta[["lka"]] - mod_pat$theta[["lka"]])
)
round(hv_ratio, 2)
stopifnot(hv_ratio[["cl"]] > 1.6, hv_ratio[["ka"]] > 13)
```

For a 70 kg subject, a 200 mg single dose (the B1371010 regimen) is simulated
with the typical healthy-volunteer model and with 200 simulated volunteers.
PKNCA's AUC0-inf from the typical-value profile must equal Dose / (CL/F).

```{r hv-sim}
mod_hv_typ <- rxode2::zeroRe(mod_hv)
hv_times <- c(0, 0.25, 0.5, 1, 1.5, 2, 3, 4, 6, 8, 12, 24, 36, 48, 72, 96, 120, 168, 240, 336)
ev_hv <- bind_rows(
  tibble(id = 1L, time = 0, evid = 1, amt = 200, cmt = "depot"),
  tibble(id = 1L, time = hv_times, evid = 0, amt = 0, cmt = "central")
) |>
  mutate(WT = 70, arm = "200 mg single dose") |>
  arrange(id, time, desc(evid))
sim_hv_typ <- rxode2::rxSolve(mod_hv_typ, events = ev_hv, returnType = "data.frame") |>
  # a one-subject solve returns no id column
  mutate(id = 1L, arm = "200 mg single dose")

nca_hv <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(sim_hv_typ |> dplyr::filter(!is.na(Cc)) |> select(id, time, Cc, arm),
                   Cc ~ time | arm + id),
  PKNCA::PKNCAdose(ev_hv |> dplyr::filter(evid == 1) |> select(id, time, amt, arm),
                   amt ~ time | arm + id),
  intervals = data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                         aucinf.obs = TRUE, half.life = TRUE)
))
hv_nca <- as.data.frame(nca_hv) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  select(PPTESTCD, PPORRES)
knitr::kable(hv_nca, digits = 2, caption = "Typical healthy volunteer, 70 kg, 200 mg single dose.")

auc_closed <- 200 / 10.1 * 1000
auc_nca <- hv_nca$PPORRES[hv_nca$PPTESTCD == "aucinf.obs"]
stopifnot(abs(auc_nca / auc_closed - 1) < 0.02)
```

```{r hv-vpc}
set.seed(20200502)
hv_cohort <- tibble(id = 1:200, WT = pmin(pmax(rnorm(200, 75, 12), 50), 110))
ev_hv_vpc <- bind_rows(
  hv_cohort |> mutate(time = 0, evid = 1, amt = 200, cmt = "depot"),
  hv_cohort |> tidyr::crossing(time = hv_times) |> mutate(evid = 0, amt = 0, cmt = "central")
) |>
  arrange(id, time, desc(evid))
sim_hv <- rxode2::rxSolve(mod_hv, events = ev_hv_vpc, returnType = "data.frame")
sim_hv |>
  dplyr::filter(time > 0) |>
  group_by(time) |>
  summarise(
    Q05 = quantile(Cc, 0.05), Q50 = quantile(Cc, 0.5), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Glasdegib Cc (ng/mL)",
       caption = "Healthy-volunteer model, 200 mg single dose, simulated 5th-95th percentiles.")
```

# Assumptions and deviations

- **IIV scale.** Table 3 gives IIV as "CV (%)" without saying how it was
  computed. The maintainers converted with the log-normal relation
  `omega^2 = log(1 + CV^2)`. For Vc/F (215.3%) this matters: reading the
  value as `100 x omega` instead would give `omega^2 = 4.64` rather than
  1.73. The paper gives no second number that could tell the two readings
  apart.
- **Residual error.** The Methods describe a proportional error on
  log-transformed concentrations with a thetarized sigma, so the patient
  model uses `lnorm()` with the Table 3 percentages as log-scale SDs (0.658
  hematologic, 0.595 solid tumor).
- **Coefficient precision.** The printed final-model equations give the
  covariate coefficients to three digits (0.173, 0.303, 0.406, 0.825, 0.653)
  where Table 3 rounds to two. The equation values are used.
- **Creatinine clearance input.** The model takes the raw Cockcroft-Gault
  CRCL (mL/min) and forms the paper's weight-standardized WNCL = CRCL x 70 /
  WT internally.
- **Marrow blasts in solid-tumor patients.** Not collected ("Not
  applicable", Table 2). The paper imputes missing continuous covariates at
  the population median, so the simulations here set 38.2% for solid-tumor
  patients, which makes the term neutral.
- **CYP3A inhibitor indicators.** The paper does not say whether they were
  time-varying. They are simple 0/1 columns, so either use works. A subject
  with both flags gets both factors, as in the published equation.
- **Healthy-volunteer model.** Table 3 footnote b states only "allometric
  body weight scaling". The exponents and the 70 kg reference are assumed to
  be the same as in the patient model. The food effect on ka was "empirically
  added ... by linear function", but its coefficient and reference food state
  are not reported, so it is omitted; ka = 0.79 1/h is the typical value in the
  unstated reference state. The paper does not say how the ketoconazole
  period of B1371010 was handled; no CYP3A inhibitor covariate is reported.
  The residual error is taken as proportional on the linear scale
  ("proportional error model", FOCEI).
- **Cohorts.** The Figure 3 cohort draws log-normal weight and CRCL around
  the Table 2 medians, truncated to the Table 2 ranges, and excludes CYP3A
  inhibitors. The healthy-volunteer weights are an assumption, as the paper
  gives none.
- No correction notice for this article was found in Europe PMC as of
  2026-09-26.
