---
title: "Recombinant Erwinia chrysanthemi asparaginase in healthy adults (Lin 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recombinant Erwinia chrysanthemi asparaginase in healthy adults (Lin 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r model}
mod <- rxode2::rxode(readModelDb("Lin_2021_asparaginaseErwiniaRecombinant"))
```

# Model and source

* Citation: `r mod$reference`
* Description: `r mod$description`
* Article: <https://doi.org/10.1002/cpdd.1002>
* Later patient model that reused this paper's absorption structure:
  `Lin_2023_asparaginaseErwiniaRecombinant` (doi:10.1111/cts.13499).

JZP-458 (marketed as Rylaze) is a recombinant *Erwinia chrysanthemi*
asparaginase produced in a *Pseudomonas fluorescens* expression platform. It is
given to patients with acute lymphoblastic leukemia (ALL) or lymphoblastic
lymphoma (LBL) who have developed hypersensitivity to *E. coli*-derived
asparaginases. The measured quantity is **serum asparaginase activity (SAA)**,
and the therapeutic target is a nadir SAA (NSAA) of at least 0.1 IU/mL.

This phase 1 model was fit jointly to intramuscular (IM) and intravenous (IV)
data, so -- unlike the later patient model -- it estimates the IM
bioavailability relative to IV and the IV volume and clearance directly.

# Population

The model was fit to 331 quantifiable SAA observations from 24 healthy adults
in the phase 1, single-center, open-label study JZP458-101 (Miami, Florida,
November 2018 to May 2019). Participants received a single dose of JZP-458 at
12.5 or 25 mg/m^2 IM, or 25 or 37.5 mg/m^2 as a 2-hour IV infusion (six per
arm), with intensive sampling to 96 h. Mean (SD) age was 38.3 (8.6) years,
weight 78.3 (9.6) kg and body surface area 1.9 (0.1) m^2; 71% were male, 83%
White and 17% Black/African American, and 96% Hispanic/Latino (Lin 2021
Table 1).

```{r population}
str(mod$population, max.level = 1)
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Lin_2021_asparaginaseErwiniaRecombinant.R` carries
an in-file comment naming its origin. They are collected here for review.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` (CL, 70 kg) | 146 mL/h | Lin 2021 Table 2 (95% CI 128.4-163.6, RSE 6.15%); Results final CL equation |
| `e_wt_cl` | 0.863 | Lin 2021 Table 2 CL row `146 x (WT/70)^0.863`; Results final CL equation |
| `lvc` (Vd) | 3030 mL | Lin 2021 Table 2 (95% CI 2655-3405, RSE 6.32%) |
| `lfdepot` (F, IM) | 0.365 | Lin 2021 Table 2 (95% CI 0.3074-0.4226, RSE 8.05%); Results "bioavailability at 36.5%" |
| `lka` (ka) | 0.0348 1/h | Lin 2021 Table 2 (95% CI 0.02942-0.04018, RSE 7.89%) |
| `lr1` (zero-order input) | 4000 IU/h | Lin 2021 Table 2 (95% CI 1569-6431, RSE 31.01%) |
| `etalcl` | 0.03502 (BSV 18.88%) | Lin 2021 Table 2 BSV% column, `log(1 + 0.1888^2)` |
| `etalvc` | 0.09783 (BSV 32.06%) | Lin 2021 Table 2 BSV% column, `log(1 + 0.3206^2)` |
| `propSd` | 0.206 | Lin 2021 Table 2 "Error model proportional 20.6%" |
| One compartment, linear elimination | n/a | Lin 2021 Results, Base model and final covariate model |
| Sequential mixed-order IM absorption | n/a | Lin 2021 Results, Base model |
| IM-only bioavailability relative to IV | n/a | Lin 2021 Results, Base model; Table 2 footnote |
| Exponential BSV, allometric weight on CL | n/a | Lin 2021 Table 2 footnotes |
| IV infusion duration 2 h | n/a | Lin 2021 Methods, Study design |

# Model structure

IV doses enter `central` directly. IM doses enter `depot` with
**sequential mixed-order absorption**: the bioavailable fraction `F * amt`
fills the depot as a zero-order input at the constant rate `R1`, while
first-order absorption `ka` drains the depot into `central` at the same time.
`R1` is the NONMEM reserved zero-order input rate for the dosing compartment,
so the input duration is `F * amt / R1`; the same encoding is used by the later
`Lin_2023_asparaginaseErwiniaRecombinant` model.

After IV dosing the terminal slope reports `kel = CL / Vd`; after IM dosing
`ka` (0.0348 /h) is smaller than `kel` (0.048 /h for a 70 kg adult), so the IM
terminal phase is absorption rate limited (flip-flop). This is why Lin 2021
reports a half-life only for the IV route.

## Intramuscular dose records must set `rate = -1`

rxode2 only consults a modelled `rate()` when the dose record asks for it. With
the default `rate = 0` the modelled rate is silently ignored and the IM dose
becomes an instantaneous bolus into the depot. The check below confirms that,
with `rate = -1`, the depot fills for exactly `F * amt / R1` hours.

```{r rate-semantics}
tv <- rxode2::zeroRe(mod)
theta <- mod$theta
cl70 <- exp(theta[["lcl"]])
v_i <- exp(theta[["lvc"]])
f_i <- exp(theta[["lfdepot"]])
ka_i <- exp(theta[["lka"]])
r1_i <- exp(theta[["lr1"]])

# Typical-value single-dose solve. route = "IM" doses the depot with the
# modelled zero-order rate; route = "IV" is a 2-hour infusion into central.
solve_one <- function(dose, route, wt = 70, times = seq(0, 240, by = 0.25),
                      im_rate = -1) {
  ev <- if (route == "IM") {
    rxode2::et(amt = dose, cmt = "depot", rate = im_rate)
  } else {
    rxode2::et(amt = dose, cmt = "central", dur = 2)
  }
  ev <- rxode2::et(ev, times, cmt = "central")
  ev <- as.data.frame(ev)
  ev$WT <- wt
  out <- rxode2::rxSolve(tv, ev, returnType = "data.frame",
                         rtol = 1e-10, atol = 1e-12)
  if (is.null(out$id)) out$id <- 1L
  out[!is.na(out$Cc), ]
}

probe_dose <- 30000   # IU; see "Dose units" below
s_im <- solve_one(probe_dose, "IM", times = seq(0, 240, by = 0.05))
s_im_bolus <- solve_one(probe_dose, "IM", times = seq(0, 240, by = 0.05),
                        im_rate = 0)
dur_expected <- f_i * probe_dose / r1_i

rate_check <- data.frame(
  Quantity = c("depot peak time (h)", "Tmax of Cc (h)", "Cmax (IU/mL)"),
  `rate = -1` = c(s_im$time[which.max(s_im$depot)],
                  s_im$time[which.max(s_im$Cc)], max(s_im$Cc)),
  `rate = 0` = c(s_im_bolus$time[which.max(s_im_bolus$depot)],
                 s_im_bolus$time[which.max(s_im_bolus$Cc)],
                 max(s_im_bolus$Cc)),
  check.names = FALSE
)
knitr::kable(rate_check, digits = 3,
             caption = "The modelled zero-order rate is active only with rate = -1.")
sprintf("expected zero-order duration F * amt / R1 = %.3f h", dur_expected)

stopifnot(
  abs(s_im$time[which.max(s_im$depot)] - dur_expected) < 0.06,
  s_im_bolus$time[which.max(s_im_bolus$depot)] == 0
)
```

# Dose units: activity units, not mass

The model's amount unit is **IU of asparaginase activity, not mg**. The
zero-order absorption rate is reported in IU/h (Lin 2021 Table 2) and
`Cc = central / vc` is IU/mL only if `central` holds IU. Lin 2021 doses in
mg/m^2 but does not state the mg-to-IU specific activity of JZP-458, so **dose
this model in IU**; no conversion factor is built into the model file.

The checks in the next three sections use only dose-independent quantities. A
reconstruction of the dose scale from the paper's own simulation table follows
them and is labelled as such.

# Validation 1: derived quantities printed in Table 2

The Table 2 footnotes print the apparent IM clearance and volume for a 70 kg
adult (`CL/F = 0.4 L/h`, `Vd/F = 8.30 L`) and the IV half-life
(`t1/2 = ln(2) / (CL / Vd) = 14.4 h`). All three follow from the `ini()` values
by arithmetic, and the half-life is also recovered from a simulated IV profile.

```{r validation-derived}
s_iv <- solve_one(probe_dose, "IV")
tail_iv <- s_iv[s_iv$time >= 24 & s_iv$time <= 96, ]
thalf_iv_sim <- log(2) /
  -stats::coef(stats::lm(log(tail_iv$Cc) ~ tail_iv$time))[[2]]

derived <- data.frame(
  Quantity = c("CL/F, 70 kg (L/h)", "Vd/F (L)", "IV t1/2 (h), arithmetic",
               "IV t1/2 (h), simulated terminal slope"),
  Model = c(cl70 / f_i / 1000, v_i / f_i / 1000, log(2) * v_i / cl70,
            thalf_iv_sim),
  Published = c(0.4, 8.30, 14.4, 14.4)
)
derived$`Difference (%)` <- 100 * (derived$Model - derived$Published) /
  derived$Published
knitr::kable(derived, digits = 3,
             caption = "Model versus the derived values printed in Lin 2021 Table 2.")

# Deterministic; the published values are rounded to 2-3 significant figures.
stopifnot(all(abs(derived$`Difference (%)`) < 1))
```

# Validation 2: exact bioavailability and weight identities

Elimination is linear, so the IM-to-IV AUC ratio at equal dose must equal `F`,
and halving body weight must raise AUC by `2^0.863` (AUC = dose / CL, with
`CL` proportional to `WT^0.863`).

```{r validation-identities}
auc_of <- function(s) {
  sum(diff(s$time) * (utils::head(s$Cc, -1) + utils::tail(s$Cc, -1)) / 2)
}
grid_long <- seq(0, 1500, by = 0.25)
auc_iv70 <- auc_of(solve_one(probe_dose, "IV", times = grid_long))
auc_im70 <- auc_of(solve_one(probe_dose, "IM", times = grid_long))
auc_iv35 <- auc_of(solve_one(probe_dose, "IV", wt = 35, times = grid_long))

id_check <- data.frame(
  Comparison = c("AUC IM / AUC IV (= F)", "AUC IV 35 kg / 70 kg (= 2^0.863)",
                 "AUC IV x CL / dose (= 1)"),
  Observed = c(auc_im70 / auc_iv70, auc_iv35 / auc_iv70,
               auc_iv70 * cl70 / probe_dose),
  Expected = c(0.365, 2^0.863, 1)
)
knitr::kable(id_check, digits = 5,
             caption = "Exposure identities reproduce F and the weight exponent.")
stopifnot(max(abs(id_check$Observed / id_check$Expected - 1)) < 1e-3)
```

# Validation 3: PKNCA on typical single-dose profiles

NCA is run on the typical-value (no IIV, no residual error) single-dose profile
for each route, at the 70 kg reference weight to which Table 2's half-life
refers. Lin 2021 reports a
half-life for the IV route only; the IM terminal phase is absorption rate
limited and reports `ln(2) / ka`.

```{r pknca}
nca_sim <- dplyr::bind_rows(
  dplyr::mutate(solve_one(probe_dose, "IV", wt = 70), treatment = "IV"),
  dplyr::mutate(solve_one(probe_dose, "IM", wt = 70), treatment = "IM")
) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, treatment, time, Cc) |>
  dplyr::mutate(Cc = pmax(Cc, 0)) |>
  dplyr::group_by(treatment) |>
  dplyr::filter(time <= time[which.max(Cc)] | Cc >= 1e-6 * max(Cc)) |>
  dplyr::ungroup()
stopifnot(all(nca_sim$time[!duplicated(nca_sim$treatment)] == 0))

dose_df <- data.frame(id = 1L, treatment = c("IV", "IM"), time = 0,
                      amt = probe_dose)
o_conc <- PKNCA::PKNCAconc(nca_sim, Cc ~ time | treatment + id)
o_dose <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)
o_data <- PKNCA::PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(start = 0, end = 240, cmax = TRUE, tmax = TRUE,
                         half.life = TRUE, auclast = TRUE)
)
nca_res <- as.data.frame(PKNCA::pk.nca(o_data))
nca_res |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "half.life", "auclast")) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::rename("Route" = treatment, "Cmax (IU/mL)" = cmax, "Tmax (h)" = tmax,
                "t1/2 (h)" = half.life, "AUC0-240 (IU*h/mL)" = auclast) |>
  knitr::kable(digits = 3,
               caption = "PKNCA on the typical single-dose profile (30000 IU probe dose).")
```

```{r pknca-compare}
hl <- nca_res |> dplyr::filter(PPTESTCD == "half.life")
stopifnot(nrow(hl) == 2L)
simulated <- data.frame(treatment = hl$treatment, PPTESTCD = "half.life",
                        PPORRES = hl$PPORRES)
reference <- data.frame(treatment = "IV", half.life = 14.4)
knitr::kable(
  nlmixr2lib::ncaComparisonTable(simulated, reference, by = "treatment",
                                 units = c(half.life = "h")),
  caption = "Typical-value IV half-life versus Lin 2021 Table 2."
)

hl_iv <- hl$PPORRES[hl$treatment == "IV"]
hl_im <- hl$PPORRES[hl$treatment == "IM"]
stopifnot(
  abs(hl_iv - 14.4) / 14.4 < 0.02,
  abs(hl_im - log(2) / ka_i) / (log(2) / ka_i) < 0.05
)
```

At the phase 1 mean weight of 78.3 kg the IV half-life would be shorter,
`r round(log(2) * v_i / (cl70 * (78.3 / 70)^0.863), 1)` h, because `CL` scales
with weight while `Vd` does not.

# Validation 4: the Table 3 simulation summary

Lin 2021 Table 3 reports, for adult and pediatric virtual populations
(1000 each, drawn from NHANES), the mean SAA and the percentage of participants
with SAA at or above 0.1 IU/mL at three times on a Monday/Wednesday/Friday
schedule (doses at 0, 48, 96, 168, 216 and 264 h): 72 h after dose 3
(t = 168 h), 48 h after dose 5 (t = 264 h) and 72 h after dose 6 (t = 336 h).

## Virtual cohorts

The NHANES records are not published, so the cohorts are approximated: 200
adults per arm (18-85 years, weight log-normal around 80 kg, height normal
around 168 cm) and 200 children per arm (2-17 years, weight and height drawn
around age-specific CDC median growth values). BSA, which sets the mg/m^2 dose,
uses the Mosteller formula; the paper does not name a BSA formula.

```{r cohort}
set.seed(2021)
n_per_arm <- 200
cdc_age <- c(2, 4, 6, 8, 10, 12, 14, 16, 17)
wt_med <- stats::approxfun(cdc_age, c(12.5, 16.3, 20.7, 25.6, 32, 40.5, 50, 58, 61),
                           rule = 2)
ht_med <- stats::approxfun(cdc_age, c(87, 102, 116, 128, 138, 150, 162, 169, 171),
                           rule = 2)
make_pop <- function(pop, n) {
  if (pop == "Adults") {
    wt <- stats::rlnorm(n, log(80), 0.22)
    ht <- stats::rnorm(n, 168, 10)
  } else {
    age <- stats::runif(n, 2, 17.99)
    wt <- wt_med(age) * stats::rlnorm(n, 0, 0.2)
    ht <- ht_med(age) * stats::rlnorm(n, 0, 0.04)
  }
  data.frame(pop = pop, WT = wt, BSA = sqrt(wt * ht / 3600))
}
regimens <- data.frame(
  regimen = c("12.5 mg/m^2 IM", "25 mg/m^2 IM", "25 mg/m^2 IV", "37.5 mg/m^2 IV"),
  mgm2 = c(12.5, 25, 25, 37.5),
  route = c("IM", "IM", "IV", "IV")
)
cohort <- tidyr::expand_grid(regimens, pop = c("Adults", "Pediatrics")) |>
  dplyr::group_by(regimen, pop) |>
  dplyr::group_modify(~ dplyr::bind_cols(.x, make_pop(.y$pop, n_per_arm)) |>
                        dplyr::select(-pop)) |>
  dplyr::ungroup() |>
  dplyr::mutate(id = dplyr::row_number())

cohort |>
  dplyr::group_by(pop) |>
  dplyr::summarise(`Median WT (kg)` = stats::median(WT),
                   `Median BSA (m^2)` = stats::median(BSA), .groups = "drop") |>
  dplyr::rename("Population" = pop) |>
  knitr::kable(digits = 2, caption = "Virtual cohort body size (per population).")
```

## Simulation at a unit dose scale

Because the model is linear in dose, the cohort is simulated with the dose
entered as `mg/m^2 x BSA` numerically (one model unit per mg). Every
concentration then scales by the same unknown specific activity `k` (IU/mg).

```{r table3-sim}
dose_times <- c(0, 48, 96, 168, 216, 264)
obs_times <- seq(0, 336, by = 2)
ev_cohort <- lapply(seq_len(nrow(cohort)), function(i) {
  amt <- cohort$mgm2[i] * cohort$BSA[i]
  dose <- if (cohort$route[i] == "IM") {
    data.frame(time = dose_times, amt = amt, cmt = "depot", rate = -1, evid = 1L)
  } else {
    data.frame(time = dose_times, amt = amt, cmt = "central", rate = amt / 2,
               evid = 1L)
  }
  obs <- data.frame(time = obs_times, amt = 0, cmt = "central", rate = 0,
                    evid = 0L)
  e <- rbind(dose, obs)
  e$id <- cohort$id[i]
  e$WT <- cohort$WT[i]
  e
}) |>
  dplyr::bind_rows() |>
  dplyr::arrange(id, time, dplyr::desc(evid))

rxode2::rxSetSeed(2021)
sim <- rxode2::rxSolve(mod, ev_cohort, returnType = "data.frame")
stopifnot(!anyNA(sim$Cc))
sim <- dplyr::left_join(sim, dplyr::select(cohort, id, regimen, pop), by = "id")
```

## Dose-scale reconstruction (not a published value)

`k` is calibrated on a single Table 3 cell -- the adult 37.5 mg/m^2 IV mean SAA
48 h after dose 5 (1.6 IU/mL, the largest adult mean and so the one with the
smallest relative rounding error). The other 11 adult mean-SAA cells and all 12
adult percentage cells are **held out**.

```{r dose-scale}
calib <- sim |>
  dplyr::filter(regimen == "37.5 mg/m^2 IV", pop == "Adults", time == 264)
k_iu_per_mg <- 1.6 / mean(calib$Cc)
sprintf("reconstructed specific activity ~ %.0f IU/mg (NOT a published value)",
        k_iu_per_mg)
sim$SAA <- sim$Cc * k_iu_per_mg
```

```{r table3-compare}
landmarks <- data.frame(
  time = c(168, 264, 336),
  landmark = c("Dose 3 72-Hour", "Dose 5 48-Hour", "Dose 6 72-Hour")
)
published <- tibble::tribble(
  ~regimen, ~pop, ~landmark, ~pct_pub, ~mean_pub,
  "12.5 mg/m^2 IM", "Adults", "Dose 3 72-Hour", 99.5, 0.3,
  "12.5 mg/m^2 IM", "Adults", "Dose 5 48-Hour", 100.0, 0.5,
  "12.5 mg/m^2 IM", "Adults", "Dose 6 72-Hour", 99.5, 0.3,
  "12.5 mg/m^2 IM", "Pediatrics", "Dose 3 72-Hour", 99.9, 0.4,
  "12.5 mg/m^2 IM", "Pediatrics", "Dose 5 48-Hour", 100.0, 0.6,
  "12.5 mg/m^2 IM", "Pediatrics", "Dose 6 72-Hour", 99.9, 0.5,
  "25 mg/m^2 IM", "Adults", "Dose 3 72-Hour", 100.0, 0.6,
  "25 mg/m^2 IM", "Adults", "Dose 5 48-Hour", 100.0, 1.0,
  "25 mg/m^2 IM", "Adults", "Dose 6 72-Hour", 100.0, 0.6,
  "25 mg/m^2 IM", "Pediatrics", "Dose 3 72-Hour", 100.0, 0.9,
  "25 mg/m^2 IM", "Pediatrics", "Dose 5 48-Hour", 100.0, 1.3,
  "25 mg/m^2 IM", "Pediatrics", "Dose 6 72-Hour", 100.0, 0.9,
  "25 mg/m^2 IV", "Adults", "Dose 3 72-Hour", 74.5, 0.4,
  "25 mg/m^2 IV", "Adults", "Dose 5 48-Hour", 95.3, 1.1,
  "25 mg/m^2 IV", "Adults", "Dose 6 72-Hour", 74.5, 0.4,
  "25 mg/m^2 IV", "Pediatrics", "Dose 3 72-Hour", 91.7, 1.2,
  "25 mg/m^2 IV", "Pediatrics", "Dose 5 48-Hour", 99.2, 2.1,
  "25 mg/m^2 IV", "Pediatrics", "Dose 6 72-Hour", 91.7, 1.3,
  "37.5 mg/m^2 IV", "Adults", "Dose 3 72-Hour", 80.9, 0.6,
  "37.5 mg/m^2 IV", "Adults", "Dose 5 48-Hour", 97.4, 1.6,
  "37.5 mg/m^2 IV", "Adults", "Dose 6 72-Hour", 80.9, 0.6,
  "37.5 mg/m^2 IV", "Pediatrics", "Dose 3 72-Hour", 94.5, 1.8,
  "37.5 mg/m^2 IV", "Pediatrics", "Dose 5 48-Hour", 99.5, 3.2,
  "37.5 mg/m^2 IV", "Pediatrics", "Dose 6 72-Hour", 94.5, 2.0
)
tab3 <- sim |>
  dplyr::inner_join(landmarks, by = "time") |>
  dplyr::group_by(regimen, pop, landmark) |>
  dplyr::summarise(pct_sim = 100 * mean(SAA >= 0.1), mean_sim = mean(SAA),
                   .groups = "drop") |>
  dplyr::inner_join(published, by = c("regimen", "pop", "landmark")) |>
  dplyr::mutate(calibration = regimen == "37.5 mg/m^2 IV" & pop == "Adults" &
                  landmark == "Dose 5 48-Hour")
stopifnot(nrow(tab3) == 24L)

tab3 |>
  dplyr::select(regimen, pop, landmark, pct_pub, pct_sim, mean_pub, mean_sim,
                calibration) |>
  dplyr::rename("Dose/route" = regimen, "Population" = pop, "Time" = landmark,
                "% >= 0.1 (Lin 2021)" = pct_pub, "% >= 0.1 (model)" = pct_sim,
                "Mean SAA (Lin 2021)" = mean_pub, "Mean SAA (model)" = mean_sim,
                "Calibration cell" = calibration) |>
  knitr::kable(digits = 2,
               caption = "Replicates Lin 2021 Table 3 (mean SAA in IU/mL).")
```

```{r table3-assert}
adults_held <- dplyr::filter(tab3, pop == "Adults", !calibration)
# Adult held-out cells. The published means are rounded to 0.1 IU/mL and the
# model means carry ~3-4% Monte-Carlo error (200 subjects per arm), as does k.
# Realised max |diff| 0.07 IU/mL; 0.2 still breaks on a mis-transcribed CL, Vd,
# F, ka or R1, each of which moves these troughs by tens of percent.
# Percentages: binomial SE at p = 0.75, n = 200 is ~3 points; 10 is > 3 SE.
stopifnot(
  max(abs(adults_held$mean_sim - adults_held$mean_pub)) < 0.2,
  stats::median(abs(adults_held$mean_sim / adults_held$mean_pub - 1)) < 0.1,
  max(abs(adults_held$pct_sim - adults_held$pct_pub)) < 10
)
```

Every adult cell is reproduced: the held-out mean SAA values agree to within
the table's 0.1 IU/mL rounding, and the IV percentages at or above target
(74.5% and 80.9% after dose 3) are matched to within Monte-Carlo error. Those
two IV percentages are sensitive to the IIV magnitudes as well as to `CL`, `Vd`
and `k`, so they also support the BSV% reading used for `etalcl` and `etalvc`.

### Scale-free checks

Two ratios in Table 3 cancel `k` entirely, so they test the structural model
without the reconstruction: the IM-to-IV mean SAA ratio at the shared
25 mg/m^2 dose, and the IV trough decay from 48 h to 72 h post dose. Each
published ratio is compared against the interval implied by the table's
0.1 IU/mL rounding.

```{r scale-free}
mean_of <- function(reg, lm) {
  tab3$mean_sim[tab3$regimen == reg & tab3$pop == "Adults" & tab3$landmark == lm]
}
ratios <- data.frame(
  Ratio = c("IM / IV, 25 mg/m^2, Dose 3 72-Hour",
            "IV 25 mg/m^2, Dose 3 72-Hour / Dose 5 48-Hour"),
  Model = c(mean_of("25 mg/m^2 IM", "Dose 3 72-Hour") /
              mean_of("25 mg/m^2 IV", "Dose 3 72-Hour"),
            mean_of("25 mg/m^2 IV", "Dose 3 72-Hour") /
              mean_of("25 mg/m^2 IV", "Dose 5 48-Hour")),
  Published = c(0.6 / 0.4, 0.4 / 1.1),
  Lower = c(0.55 / 0.45, 0.35 / 1.15),
  Upper = c(0.65 / 0.35, 0.45 / 1.05)
)
knitr::kable(ratios, digits = 3,
             caption = "Dose-scale-free ratios versus Lin 2021 Table 3 (adults).")
stopifnot(all(ratios$Model > ratios$Lower & ratios$Model < ratios$Upper))
```

### Pediatric cells

The pediatric cells are reproduced only qualitatively. The model gives the
paper's direction and order of magnitude -- children reach higher SAA than
adults on the same mg/m^2 dose, most strongly on the IV route, because `Vd` has
no body-size covariate while the dose scales with BSA and `CL` with
`WT^0.863` -- but it overshoots the published pediatric means by roughly 5-25%
on IM and 20-35% on IV. The pediatric cohort here is built from CDC median
growth values, not from the NHANES records the paper resampled, and pediatric
exposure in this model is sensitive to the weight-to-BSA relationship. The gap
is therefore attributed to the cohort, not to the model, but it is recorded in
the Errata rather than gated tightly.

```{r pediatric-assert}
ped <- dplyr::filter(tab3, pop == "Pediatrics")
ped_iv_ratio <- ped$mean_sim[grepl("IV", ped$regimen)] /
  tab3$mean_sim[tab3$pop == "Adults" & grepl("IV", tab3$regimen)]
stopifnot(
  # Children above adults on IV: the model gives ~2.5-4x, far from 1.
  all(ped_iv_ratio > 1.5),
  # Pediatric means within a factor of 1.6 of the published values.
  all(ped$mean_sim / ped$mean_pub > 1 / 1.6 & ped$mean_sim / ped$mean_pub < 1.6),
  all(ped$pct_sim > 85)
)
```

## Figure 4: median SAA with 90% prediction interval

```{r figure4, fig.width = 8, fig.height = 5}
fig4 <- sim |>
  dplyr::group_by(regimen, pop, time) |>
  dplyr::summarise(lo = stats::quantile(SAA, 0.05), med = stats::median(SAA),
                   hi = stats::quantile(SAA, 0.95), .groups = "drop") |>
  dplyr::mutate(panel = paste(regimen, "-", pop),
                lo = pmax(lo, 1e-3))

ggplot2::ggplot(fig4, ggplot2::aes(time)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi), alpha = 0.3,
                       fill = "steelblue") +
  ggplot2::geom_line(ggplot2::aes(y = med), colour = "steelblue",
                     linetype = "dashed") +
  ggplot2::geom_hline(yintercept = 0.1, linetype = "dashed") +
  ggplot2::scale_y_log10(limits = c(0.005, 30)) +
  ggplot2::scale_x_continuous(breaks = seq(0, 288, by = 72)) +
  ggplot2::facet_wrap(~panel, nrow = 2) +
  ggplot2::labs(x = "Time (hours)", y = "SAA (IU/mL)",
                caption = paste("Replicates Figure 4 of Lin 2021, using the",
                                "reconstructed dose scale.")) +
  ggplot2::theme_bw(base_size = 8)
```

The reconstructed specific activity can be compared with the independent
reconstruction in the `Lin_2023_asparaginaseErwiniaRecombinant` article, which
calibrated the later patient model to a figure of that paper and landed near
580 IU/mg. The two agree to within the precision either can claim.

```{r dose-scale-assert}
stopifnot(k_iu_per_mg > 450, k_iu_per_mg < 900)
```

# Assumptions and deviations / Errata

1. **The model's amount unit is IU, and no mg-to-IU conversion is published.**
   The zero-order absorption rate is reported in IU/h and `Cc = central / vc`
   is IU/mL only if the dose is in IU. Lin 2021 doses in mg/m^2 but does not
   state the specific activity of JZP-458. **Dose this model in IU.** No
   conversion factor is present in the model file.

2. **The dose scale used for the Table 3 and Figure 4 replications is
   reconstructed, not published.** It is calibrated on one Table 3 cell and
   implies about `r round(k_iu_per_mg, -1)` IU/mg. It then reproduces the 11
   held-out adult mean-SAA cells and 12 adult percentage cells, and two
   dose-scale-free ratios confirm the structural model independently of it.

3. **BSV% read as a coefficient of variation.** Table 2 prints `BSV%` of
   18.88 (CL) and 32.06 (Vd) and states that BSV was exponential. These are
   converted with `omega^2 = log(1 + CV^2)` (0.03502 and 0.09783). Reading the
   column as `100 * omega` instead would give 0.03565 and 0.1028; at these
   magnitudes the two readings differ by under 5% in variance and cannot be
   separated by the published results. No CL-Vd covariance is reported, so the
   etas are uncorrelated.

4. **The weight exponent is encoded as estimated.** 0.863 appears only inside
   the CL row of Table 2 and in the printed final CL equation, with no separate
   RSE or CI. It is not a conventional value such as 0.75 and the Methods
   describe weight as a covariate retained by the OFV-drop criterion, so it is
   not wrapped in `fixed()`.

5. **Absorption structure inferred from prose plus the parameter set.** Lin
   2021 describes "a sequential mixed order absorption function" estimating
   "both zero- and first-order absorption rate constant parameters" and reports
   a zero-order rate in IU/h, but prints no absorption equation. The encoding
   -- the bioavailable dose filling the depot at `R1` while `ka` drains it --
   uses exactly the reported `ka`, `R1` and `F`, matches NONMEM's reserved `R1`
   semantics, and is the same encoding as the later Lin 2023 patient model.

6. **IV infusion duration.** The IV doses are 2-hour infusions (Lin 2021
   Methods, Study design), encoded as `dur = 2` / `rate = amt / 2` on records
   into `central`. The bioavailability `F` applies to the IM depot only.

7. **Virtual cohorts approximate NHANES.** The paper resampled 2000 NHANES
   subjects (1000 adults, 1000 aged 2-17 years); those records are not
   published. The cohorts here (200 per arm and population) use simple
   distributional approximations and the Mosteller BSA formula, which the paper
   does not name. The pediatric cohort is the likely cause of the 5-35%
   overshoot of the published pediatric means (see "Pediatric cells").

8. **`Cc` is the individual prediction, without residual error.** Table 3's
   percentages may include residual error in the paper's simulation; the
   adult percentages are matched within Monte-Carlo error without it.

9. **Table 2 parameter differences from the Lin 2023 patient model.** Lin 2023
   reports `ka` 0.0369 1/h, zero-order rate 1810 IU/h and `F` 0.359 as fixed
   from "a PopPK model developed with intensive phase I data". Those values
   differ from this paper's Table 2 (0.0348 1/h, 4000 IU/h, 0.365). This model
   uses the values printed here.

10. **Screened-but-not-retained covariates are documented, not modelled.** BSA
    (significant but not chosen over weight), age, sex and race are recorded in
    the model file's `covariatesDataExcluded` list.
