---
title: "Pneumocystis treatment QSP model (Liu 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pneumocystis treatment QSP model (Liu 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Liu 2018 builds three things and then composes them: a three-compartment PK
module, instantiated once per drug; a two-stage *Pneumocystis* life-cycle PD
module; and four integrated QSP models, one per drug. The extraction mirrors
that structure, so the paper contributes **five** model files to `nlmixr2lib`
and this one vignette validates all of them.

```{r models}
qsp_models <- c(
  anidulafungin = "Liu_2018_anidulafungin_mouse_qsp",
  caspofungin = "Liu_2018_caspofungin_mouse_qsp",
  micafungin = "Liu_2018_micafungin_mouse_qsp",
  sulfamethoxazole = "Liu_2018_sulfamethoxazole_mouse_qsp"
)
pd_model <- "Liu_2018_pneumocystis_mouse_qsp"

# readModelDb() returns the model FUNCTION; rxode2::rxode() resolves it to the
# rxUi whose metadata fields are readable.
ui <- lapply(c(qsp_models, pneumocystis = pd_model), \(n) rxode2::rxode(readModelDb(n)))
```

* Citation: `r ui[["anidulafungin"]]$reference[1]`
* Article: <https://doi.org/10.1186/s12918-018-0603-9>
* PubMed Central: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6050661/>

```{r model-table, echo = FALSE}
tibble::tibble(
  Model = names(ui),
  File = c(unname(qsp_models), pd_model),
  Role = c(
    "Integrated QSP: anidulafungin PK + Pneumocystis PD",
    "Integrated QSP: caspofungin PK + Pneumocystis PD",
    "Integrated QSP: micafungin PK + Pneumocystis PD",
    "Integrated QSP: sulfamethoxazole (TMP-SMX proxy) PK + Pneumocystis PD, with the 7-day effect delay",
    "PD module alone: untreated Pneumocystis natural history"
  )
) |>
  dplyr::rename("nlmixr2lib model" = File) |>
  knitr::kable(caption = "The five model files contributed by Liu 2018.")
```

# Population

The model was constrained with data from mice. Liu 2018's own experiments used
**6-week-old male C3H/HeN mice** (Charles River) that were immunosuppressed and
infected with *Pneumocystis murina*; organism burden was read out either by
RT-qPCR of *Pneumocystis* mitochondrial large-subunit rRNA (total nuclei, which
cannot distinguish the two life-cycle stages) or by microscopic quantification,
using cresyl echt violet to stain asci specifically and a rapid Wright-Giemsa
stain to count nuclei of all stages. The nuclei-count time course of Figure 3b
has n = 2 or 3 mice per time point.

The PK modules were **not** fitted to new data. Liu 2018 digitised published
murine plasma-concentration-versus-time figures with `labnotes` and tuned each
parameter set by hand: anidulafungin from Gumbo 2006 (construction) and Andes
2008 (validation); caspofungin and micafungin from Andes 2010, with Hajdu 1997
as additional caspofungin validation; sulfamethoxazole from Misiek 1985.
Organism burdens under treatment came from Cushion 2010, the echinocandin
depletion study that motivated the two-stage life-cycle structure.

The TMP:SMX ratio was fixed at 1:5 in the constraining data, so Liu 2018 "used
the level of SMX as a reasonable proxy for this drug combination" -- the
sulfamethoxazole model therefore carries a single drug, not two.

The same information is available programmatically:

```{r population}
str(ui[["anidulafungin"]]$population)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry. The table below collects them in one place.

| Equation / parameter | Value | Source location |
|---|---|---|
| `d/dt(depot) = -RAP*Kabs*depot` | n/a | Table 1 "Absorptive Compartment"; Methods equation (a) |
| `d/dt(central) = Kabs*depot - (KPT+KdP)*central + KTP*peripheral1` | n/a | Table 1 "Plasma Compartment"; Methods equation (b) |
| `d/dt(peripheral1) = RTP*(-KTP*peripheral1 + KPT*central) - KdT*peripheral1` | n/a | Table 1 "Peripheral Tissue Compartment"; Methods equation (c) |
| `lka` (Kabs) | 5 / 5 / 0.75 / 5 per h | Table 1, row `Kabs` |
| `lk12` (KPT) | 1.5 / 5 / 0.6 / 0.17 per h | Table 1, row `KPT` |
| `lk21` (KTP) | 5 / 2 / 1.8 / 5 per h | Table 1, row `KTP` |
| `lkel` (KdP) | 0.035 / 0.18 / 0.06 / 0.2 per h | Table 1, row `KdP` |
| `lkdt` (KdT) | 0.035 / 0.18 / 0.06 / 0.2 per h | Table 1, row `KdT` |
| `rap` (RAP) | 3 / 0.1 / 1 / 3 | Table 1, row `RAP` |
| `rtp` (RTP) | 0.2 / 1 / 0.4 / 0.01 | Table 1, row `RTP` |
| Initial AC concentration per dose | Table of 6 doses x 3 echinocandins | Table 3; SMX from the Table 3 footnote ("For SMX, 500 and 550 (ug/ml) were used for applied dosages of 200 and 250 (mg/kg) respectively") |
| `d/dt(trophic) = KsTro*trophic - KdTro*trophic^2 - KTA*trophic + KAT*asci` | n/a | Table 4 "trophic Form"; Methods equation (d) |
| `d/dt(asci) = KTA*trophic - KAT*asci - KdAsci*asci` | n/a | Table 4 "asci"; Methods equation (e) |
| `kstro` (KsTro) | 1 per day | Table 4, "Basal Parameter Values" |
| `kdtro` (KdTro) | 1e-7 per day | Table 4, "Basal Parameter Values" |
| `kat` (KAT) | 0.1 per day | Table 4, "Basal Parameter Values" |
| `kta` (KTA) | 0.1 per day | Table 4, "Basal Parameter Values" |
| `kdasci` (KdAsci) | 2e-12 per day | Table 4, "Basal Parameter Values" |
| `vdasci = kdAsci + ME*E/(E+Ec50)` (echinocandin) | n/a | Table 5, "Echinocandin effect on asci death" |
| `vta = kTA*(1 - E/(E+Ec50))` (echinocandin) | n/a | Table 5, "Echinocandin effect on asci Formation" |
| `ec50`, `emax`, `hill` (echinocandin) | 0.039 / 0.0007 / 0.04 ug/mL; 0.42 / 0.45 / 0.1 per day; n = 1 | Table 5, echinocandin parameter table |
| `vdasci = kdAsci + MEAsci*E/(E+Ec50)` (SMX) | n/a | Table 5, "TMP/SMX effect on asci death" |
| `vstro = ks*(1 - E/(E+Ec50))` (SMX) | n/a | Table 5, "TMP/SMX effect on trophic proliferation"; Methods |
| `vdtro = kdTro*(1 + METro*E/(E+Ec50))` (SMX) | n/a | Table 5, "TMP/SMX effect on trophic death"; Methods |
| `smxeff = central(t - tau)` | n/a | Table 5, "Delay in SMX effect" |
| `ec50`, `hill`, `emax_asci`, `emax_trophic`, `tau` (SMX) | 0.2 ug/mL; n = 2; 0.75 per day; 650; 7 days | Table 5, TMP/SMX parameter table |
| `trophic0`, `asci0` | 1 organism each | **Not reported.** Back-solved from the paper's described untreated time course; see "Assumptions and deviations" |
| PK-to-PD time-base conversion (x 24) | n/a | Derived: Table 1 is in 1/h, Table 4 in 1/day, Table 5's tau in days |
| Dosing schedule | 3 doses/week for 3 weeks | Results, "Following the experimental dosing regimen reported by Cushion et al., each drug was elevated three times a week for 3 weeks" |

# Part 1 -- the PD module

## Closed-form steady state (exact)

The drug-free system has an analytic non-trivial fixed point. Setting
`d/dt(asci) = 0` gives `asci = KTA*trophic/(KAT + KdAsci)`; substituting into
`d/dt(trophic) = 0` gives

```
trophic_ss = (KsTro - KTA + KAT*KTA/(KAT + KdAsci)) / KdTro
```

Because `KdAsci` is twelve orders of magnitude below `KAT`, the two
transformation terms cancel to working precision and `trophic_ss` collapses to
`KsTro / KdTro = 1 / 1e-7 = 1e7` -- exactly the "steady state of about 10^7"
Liu 2018 reports. This is a deterministic identity, so it is asserted tightly.

```{r pd-steady-state}
pd_par <- ui[["pneumocystis"]]$theta
tro_ss <- with(
  as.list(pd_par),
  (kstro - kta + kat * kta / (kat + kdasci)) / kdtro
)
asci_ss <- with(as.list(pd_par), kta * tro_ss / (kat + kdasci))

# Integrate far past equilibration and compare against the closed form.
ev_pd_long <- data.frame(
  id = 1L, time = seq(0, 400, by = 1),
  amt = NA_real_, evid = 0L, cmt = "trophic"
)
sim_pd_long <- rxode2::rxSolve(
  readModelDb(pd_model),
  events = ev_pd_long,
  atol = 1e-12, rtol = 1e-10,
  useLinCmt = FALSE
) |>
  as.data.frame()

tro_num <- tail(sim_pd_long$trophic, 1)
asci_num <- tail(sim_pd_long$asci, 1)

tibble::tibble(
  Quantity = c("trophic", "asci"),
  `Closed form` = c(tro_ss, asci_ss),
  `Numeric (day 400)` = c(tro_num, asci_num),
  `Relative difference` = c(abs(tro_num - tro_ss) / tro_ss, abs(asci_num - asci_ss) / asci_ss)
) |>
  knitr::kable(digits = c(0, 0, 0, 12), caption = "Drug-free steady state: solver vs. closed form.")

# Deterministic identity -- no cohort, no RNG -- so a tight bound is correct.
stopifnot(
  abs(tro_num - tro_ss) / tro_ss < 1e-6,
  abs(asci_num - asci_ss) / asci_ss < 1e-6,
  abs(tro_ss - 1e7) / 1e7 < 1e-9
)
```

## Untreated accumulation (Figure 3b, and the pre-treatment phase of Figure 4c)

Liu 2018 describes the untreated time course in words: "The initial growth of
the organism was very slow within the first two weeks, however, starting from
the third week, an exponential growth of *Pneumocystis* was observed which
peaked at the end of the fifth week", and "At about 35 days, the levels of both
trophic forms and asci reached a steady state of about 10^7". Those three
landmarks are what the day-0 inoculum was back-solved against (see
"Assumptions and deviations").

```{r figure-3b}
ev_pd <- data.frame(
  id = 1L, time = seq(0, 60, by = 0.25),
  amt = NA_real_, evid = 0L, cmt = "trophic"
)
sim_pd <- rxode2::rxSolve(
  readModelDb(pd_model),
  events = ev_pd,
  atol = 1e-12, rtol = 1e-10,
  useLinCmt = FALSE
) |>
  as.data.frame()

total_nuclei <- sim_pd$trophic + sim_pd$asci
sim_pd$normalised_total <- total_nuclei / max(total_nuclei)

sim_pd |>
  dplyr::select(time, trophic, asci) |>
  tidyr::pivot_longer(-time, names_to = "stage", values_to = "organisms") |>
  ggplot(aes(time, organisms, colour = stage)) +
  geom_line(linewidth = 0.9) +
  scale_y_log10(limits = c(1, 3e7)) +
  geom_vline(xintercept = 35, linetype = "dashed") +
  labs(
    x = "Time (day)", y = "Organisms per lung (log10 scale)",
    colour = "Life stage",
    title = "Untreated Pneumocystis accumulation",
    caption = paste(
      "Replicates the drug-free dynamics of Figure 3b and the pre-treatment",
      "phase of Figure 4c of Liu 2018. Dashed line: day 35."
    )
  )
```

```{r figure-3b-landmarks}
at <- function(d) sim_pd$normalised_total[which.min(abs(sim_pd$time - d))]
frac_35 <- sim_pd$trophic[which.min(abs(sim_pd$time - 35))] / tro_ss
t_half <- sim_pd$time[which(sim_pd$trophic >= 0.5 * tro_ss)[1]]

tibble::tibble(
  Landmark = c(
    "Normalised total nuclei at day 14 (paper: 'very slow' growth)",
    "Normalised total nuclei at day 21 (paper: exponential phase begins)",
    "Trophic burden at day 35 as a fraction of the plateau (paper: 'about 35 days ... steady state')",
    "Day at which the trophic burden crosses half the plateau"
  ),
  Value = c(at(14), at(21), frac_35, t_half)
) |>
  knitr::kable(digits = 3, caption = "Figure 3b landmarks reproduced by the PD module.")

# Deterministic run (no IIV in this model), so exact bounds are appropriate.
stopifnot(
  at(14) < 0.05, # still in the slow phase two weeks in
  at(21) > at(14), # exponential growth has started by week three
  frac_35 > 0.95, # plateau essentially reached by day 35
  t_half > 14, t_half < 25 # the rise is centred in weeks three and four
)
```

## Asci repopulation after echinocandin withdrawal (Figure 3a)

Figure 3a starts "from an initial state with a high level of trophic forms and
a low level of asci" and shows that "it takes several weeks for asci to
repopulate". Rather than invent that initial state, it is taken from the
anidulafungin QSP model's own day-56 output (built in Part 3 below) and the
drug-free PD module is then integrated forward.

```{r figure-3a}
dose_days <- c(35, 37, 39, 42, 44, 46, 49, 51, 53)

make_qsp_events <- function(ac0, dose_days, tmax = 56, by = 0.25) {
  obs <- data.frame(
    id = 1L, time = seq(0, tmax, by = by),
    amt = NA_real_, evid = 0L, cmt = "trophic"
  )
  if (is.null(ac0)) {
    return(obs[order(obs$time), ])
  }
  dose <- data.frame(
    id = 1L, time = dose_days, amt = ac0, evid = 1L, cmt = "depot"
  )
  out <- dplyr::bind_rows(dose, obs)
  out[order(out$time, -out$evid), ]
}

solve_qsp <- function(model, events) {
  rxode2::rxSolve(
    readModelDb(model),
    events = events,
    atol = 1e-12, rtol = 1e-10, maxsteps = 1e6,
    useLinCmt = FALSE
  ) |>
    as.data.frame()
}

# Anidulafungin 1 mg/kg i.p. => initial AC concentration 10 ug/mL (Table 3).
sim_anid <- solve_qsp(qsp_models[["anidulafungin"]], make_qsp_events(10, dose_days))
post_treatment <- c(
  trophic = tail(sim_anid$trophic, 1),
  asci = tail(sim_anid$asci, 1)
)

ev_recover <- data.frame(
  id = 1L, time = seq(0, 90, by = 0.25),
  amt = NA_real_, evid = 0L, cmt = "trophic"
)
sim_recover <- rxode2::rxSolve(
  readModelDb(pd_model),
  events = ev_recover,
  inits = post_treatment,
  atol = 1e-12, rtol = 1e-10,
  useLinCmt = FALSE
) |>
  as.data.frame()

sim_recover |>
  dplyr::select(time, trophic, asci) |>
  tidyr::pivot_longer(-time, names_to = "stage", values_to = "organisms") |>
  ggplot(aes(time, organisms, colour = stage)) +
  geom_line(linewidth = 0.9) +
  scale_y_log10() +
  labs(
    x = "Days after withdrawal of anidulafungin", y = "Organisms per lung (log10 scale)",
    colour = "Life stage",
    title = "Asci repopulate over several weeks once the echinocandin stops",
    caption = "Replicates Figure 3a of Liu 2018."
  )
```

```{r figure-3a-check}
# Time for the asci to recover to within 0.3 log10 of their drug-free plateau.
t_recover <- sim_recover$time[which(sim_recover$asci >= 10^(log10(asci_ss) - 0.3))[1]]
cat(sprintf(
  "Asci start %.2f log10 below the drug-free plateau and recover to within 0.3 log10 of it after %.1f days (%.1f weeks).\n",
  log10(asci_ss) - log10(post_treatment[["asci"]]), t_recover, t_recover / 7
))
# The paper's claim is qualitative -- "several weeks". Bracket it loosely
# enough to be a real test of the KTA / KAT timescale and no tighter.
stopifnot(t_recover > 7, t_recover < 70)
```

## Population of PD models (Figures 3c and 3d)

"To recapture these experimentally observed distributions, we constructed a
population of PD models with parameter values selected from a uniform
distribution that covers 70-130% of the basal values (Table 4). ... In order to
incorporate variability, all model parameters are changed independently."

The five PD rate constants are therefore resampled independently on
`U(0.7 x basal, 1.3 x basal)`. These draws come from R's RNG, not rxode2's, so
the cohort is reproducible across machines and thread counts.

```{r figure-3cd}
set.seed(20260918)
n_pop <- 200L # cohort cap is 200 per arm
pd_names <- c("kstro", "kdtro", "kta", "kat", "kdasci")
basal <- pd_par[pd_names]

pop <- as.data.frame(lapply(basal, \(b) runif(n_pop, 0.7 * b, 1.3 * b)))
pop$trophic0 <- pd_par[["trophic0"]]
pop$asci0 <- pd_par[["asci0"]]
pop$id <- seq_len(n_pop)

ev_pop <- tidyr::expand_grid(id = pop$id, time = c(0, 60)) |>
  dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "trophic") |>
  as.data.frame()

sim_pop <- rxode2::rxSolve(
  readModelDb(pd_model),
  params = pop,
  events = ev_pop,
  atol = 1e-12, rtol = 1e-10,
  useLinCmt = FALSE
) |>
  as.data.frame() |>
  dplyr::filter(time == 60)

sim_pop |>
  dplyr::select(id, log10_trophic, log10_asci) |>
  tidyr::pivot_longer(-id, names_to = "stage", values_to = "log10_organisms") |>
  dplyr::mutate(stage = dplyr::recode(stage,
    log10_trophic = "trophic form",
    log10_asci = "asci"
  )) |>
  ggplot(aes(log10_organisms)) +
  geom_histogram(bins = 25) +
  facet_wrap(~stage) +
  labs(
    x = "log10 organisms per lung at day 60", y = "Simulated mice",
    title = "Simulated burden distributions under 70-130% parameter resampling",
    caption = "Replicates Figures 3c and 3d of Liu 2018."
  )
```

```{r figure-3cd-table}
# Liu 2018 Results: 'the simulated distributions of the trophic forms (Fig. 3c)
# and asci (Fig. 3d) are consistent to the observed levels of the fungi
# (7.62 +/- 0.17 for trophic forms and 7.79 +/- 0.13 for asci) [33]'.
burden_summary <- tibble::tibble(
  Stage = c("trophic form", "asci"),
  `Simulated median (log10)` = c(median(sim_pop$log10_trophic), median(sim_pop$log10_asci)),
  `Simulated range (log10)` = c(
    sprintf("%.2f - %.2f", min(sim_pop$log10_trophic), max(sim_pop$log10_trophic)),
    sprintf("%.2f - %.2f", min(sim_pop$log10_asci), max(sim_pop$log10_asci))
  ),
  `Observed, Cushion 2010 (log10)` = c("7.62 +/- 0.17", "7.79 +/- 0.13")
)
knitr::kable(burden_summary, digits = 2, caption = "Simulated vs. observed steady-state burdens.")

# The simulated plateau is KsTro / KdTro exactly, so a 70-130% resample can
# reach at most log10(1.3 / (0.7e-7)) = 7.27. The observed means of 7.62 and
# 7.79 therefore sit ABOVE anything this parameterisation can produce -- a
# reproducible deviation from the paper's consistency claim, recorded under
# 'Assumptions and deviations' rather than papered over. What is asserted here
# is the arithmetic the model does commit to.
stopifnot(
  # Centre of the resampled distribution is the basal plateau, log10(1e7) = 7.
  abs(median(sim_pop$log10_trophic) - 7) < 0.15,
  # And every draw lies inside the algebraic envelope of the resampling.
  all(sim_pop$log10_trophic >= log10(0.7 / 1.3e-7) - 0.01),
  all(sim_pop$log10_trophic <= log10(1.3 / 0.7e-7) + 0.01)
)
```

# Part 2 -- the PK module

## Closed-form plasma AUC (exact)

Integrating the three PK equations from zero to infinity turns them into three
linear algebraic equations in the three areas. Solving them gives an exact
plasma AUC for any dose placed in the administration compartment:

```
AUC_plasma = (AC0 / RAP) / ( (KPT + KdP) - KPT*KTP*RTP / (RTP*KTP + KdT) )
```

Both scaling factors appear where the paper's asymmetric structure puts them:
`RAP` divides the administered concentration (only a fraction `1/RAP` of it
reaches plasma) and `RTP` enters the tissue-recirculation term only. This is an
algebraic identity, so agreement with the solver should be at machine
precision -- which makes it a sharp check that the ODEs were transcribed
correctly, including the asymmetries.

```{r pk-auc-closed-form}
# Table 3 row 1: the highest tabulated i.p. dose per echinocandin, and the
# 200 mg/kg oral sulfamethoxazole dose from the Table 3 footnote.
ac0_single <- c(
  anidulafungin = 90, # 10 mg/kg i.p.
  caspofungin = 12, # 10 mg/kg i.p.
  micafungin = 25, # 10 mg/kg i.p.
  sulfamethoxazole = 500 # 200 mg/kg p.o.
)

auc_closed <- function(model) {
  p <- as.list(rxode2::rxode(readModelDb(model))$theta)
  ka <- exp(p$lka)
  k12 <- exp(p$lk12)
  k21 <- exp(p$lk21)
  kel <- exp(p$lkel)
  kdt <- exp(p$lkdt)
  1 / ((k12 + kel) - k12 * k21 * p$rtp / (p$rtp * k21 + kdt)) / p$rap
}

# Single i.p. / p.o. dose, observed on a fine grid over 10 days. Observation
# rows sit on the `central` ODE state so that the algebraic observable Cc is
# returned at every one of them.
make_pk_events <- function(ac0, tmax = 10, n = 4001L) {
  dose <- data.frame(id = 1L, time = 0, amt = ac0, evid = 1L, cmt = "depot")
  obs <- data.frame(
    id = 1L, time = seq(0, tmax, length.out = n),
    amt = NA_real_, evid = 0L, cmt = "central"
  )
  out <- dplyr::bind_rows(dose, obs)
  out[order(out$time, -out$evid), ]
}

sim_pk <- lapply(names(qsp_models), function(d) {
  s <- solve_qsp(qsp_models[[d]], make_pk_events(ac0_single[[d]]))
  s$drug <- d
  # rxSolve omits `id` for a single-subject event table.
  s$id <- 1L
  s
}) |>
  dplyr::bind_rows() |>
  # Report time in hours: Table 1 is in 1/h and Figure 5 is plotted in hours.
  dplyr::mutate(time_h = time * 24)

stopifnot(all(sim_pk$Cc >= 0))
```

```{r figure-5}
sim_pk |>
  dplyr::filter(time_h <= 48) |>
  ggplot(aes(time_h, Cc)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~drug, scales = "free_y") +
  labs(
    x = "Time (h)", y = "Plasma concentration (ug/mL)",
    title = "Single-dose plasma profiles from the four PK modules",
    caption = paste(
      "Replicates the black construction curves of Figure 5 of Liu 2018",
      "(anidulafungin, caspofungin, micafungin 10 mg/kg i.p.;",
      "sulfamethoxazole 200 mg/kg p.o.)."
    )
  )
```

The module also reproduces the paper's i.v.-versus-i.p. contrast (Figure 5a,
blue): "in order to mimic i.v. injection, we elevated the initial level of the
drug in the plasma compartment", which is a dose into `central` rather than
`depot`. The two routes share the same disposition parameters, so an i.v. dose
of `AC0/RAP` must give the same plasma AUC as an i.p. dose of `AC0` -- another
exact consequence of the structure.

```{r figure-5a-iv}
ev_iv <- dplyr::bind_rows(
  data.frame(id = 1L, time = 0, amt = 90 / 3, evid = 1L, cmt = "central"),
  data.frame(
    id = 1L, time = seq(0, 10, length.out = 4001L),
    amt = NA_real_, evid = 0L, cmt = "central"
  )
)
ev_iv <- ev_iv[order(ev_iv$time, -ev_iv$evid), ]
sim_iv <- solve_qsp(qsp_models[["anidulafungin"]], ev_iv)

dplyr::bind_rows(
  dplyr::mutate(sim_iv, route = "i.v. 30 ug/mL into plasma"),
  dplyr::mutate(
    dplyr::filter(sim_pk, drug == "anidulafungin"),
    route = "i.p. 90 ug/mL into the AC"
  )
) |>
  dplyr::filter(time * 24 <= 36) |>
  ggplot(aes(time * 24, Cc, colour = route)) +
  geom_line(linewidth = 0.9) +
  labs(
    x = "Time (h)", y = "Plasma anidulafungin (ug/mL)", colour = NULL,
    title = "Anidulafungin: i.v. versus i.p. administration",
    caption = "Replicates the route contrast of Figure 5a of Liu 2018."
  ) +
  theme(legend.position = "bottom")
```

## PKNCA on the simulated plasma profiles

Liu 2018 reports no NCA table -- it reports SSE against digitised profiles
instead, and argues in the Discussion that the temporal predictions are more
informative than "traditional pharmacokinetic indexes such as area under the
curve". The NCA below is therefore not a comparison against published NCA
values; it exists to turn the exact AUC identity above into an independent,
tool-computed check of the same quantity, and to put interpretable Cmax / Tmax
/ half-life numbers on each module.

```{r pknca}
# Only `!is.na(Cc)` -- filtering on time or Cc would drop the time-zero row
# that PKNCA needs to anchor AUC.
sim_nca <- sim_pk |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time = time_h, Cc, drug)

sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, drug) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, drug, time, .keep_all = TRUE) |>
  dplyr::arrange(drug, id, time)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(sim_nca), Cc ~ time | drug + id)

dose_df <- sim_pk |>
  dplyr::distinct(drug, id) |>
  dplyr::mutate(time = 0, amt = ac0_single[drug]) |>
  as.data.frame()
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | drug + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r pknca-table}
nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(drug, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

nca_wide$auc_closed <- vapply(
  nca_wide$drug,
  \(d) auc_closed(qsp_models[[d]]) * ac0_single[[d]],
  numeric(1)
)
nca_wide$auc_pct_diff <- 100 * (nca_wide$aucinf.obs - nca_wide$auc_closed) / nca_wide$auc_closed

nca_wide |>
  dplyr::select(drug, cmax, tmax, half.life, aucinf.obs, auc_closed, auc_pct_diff) |>
  dplyr::rename(
    "Drug" = drug,
    "Cmax (ug/mL)" = cmax,
    "Tmax (h)" = tmax,
    "t1/2 (h)" = half.life,
    "PKNCA AUCinf (ug*h/mL)" = aucinf.obs,
    "Closed-form AUC (ug*h/mL)" = auc_closed,
    "Difference (%)" = auc_pct_diff
  ) |>
  knitr::kable(
    digits = c(0, 2, 2, 2, 1, 1, 4),
    caption = paste(
      "PKNCA parameters of the four PK modules after a single dose, with",
      "AUCinf checked against the closed-form identity."
    )
  )

# Deterministic models and an algebraic target: the agreement is numerical
# error only, so the bound is tight. A transcription error in any of ka, k12,
# k21, kel, kdt, rap or rtp moves this by whole percent or more.
stopifnot(
  nrow(nca_wide) == 4L,
  all(!is.na(nca_wide$aucinf.obs)),
  max(abs(nca_wide$auc_pct_diff)) < 0.5
)
```

## Multiple dosing three times a week (Figure 6)

```{r figure-6}
mw_days <- c(0, 2, 4, 7, 9, 11, 14, 16, 18)
echino_doses <- list(
  anidulafungin = c(`10 mg/kg` = 90, `1 mg/kg` = 10, `0.1 mg/kg` = 1),
  caspofungin = c(`10 mg/kg` = 12, `1 mg/kg` = 2.5, `0.1 mg/kg` = 0.8),
  micafungin = c(`10 mg/kg` = 25, `1 mg/kg` = 5, `0.1 mg/kg` = 1)
)

sim_fig6 <- lapply(names(echino_doses), function(d) {
  lapply(names(echino_doses[[d]]), function(lv) {
    ev <- make_qsp_events(echino_doses[[d]][[lv]], mw_days, tmax = 21, by = 0.02)
    s <- solve_qsp(qsp_models[[d]], ev)
    s$drug <- d
    s$dose <- lv
    s
  }) |> dplyr::bind_rows()
}) |>
  dplyr::bind_rows()

sim_fig6 |>
  dplyr::filter(time <= 7) |>
  ggplot(aes(time, Cc, colour = dose)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~drug, scales = "free_y") +
  labs(
    x = "Time (day)", y = "Plasma concentration (ug/mL)", colour = "Dose",
    title = "Echinocandin plasma levels over one week of 3-doses-per-week treatment",
    caption = "Replicates panels a-c of Figure 6 of Liu 2018 (first week shown)."
  ) +
  theme(legend.position = "bottom")
```

```{r figure-6-smx}
ev_smx3w <- make_qsp_events(500, mw_days, tmax = 21, by = 0.02)
sim_smx3w <- solve_qsp(qsp_models[["sulfamethoxazole"]], ev_smx3w)

ggplot(sim_smx3w, aes(time, Cc)) +
  geom_line(linewidth = 0.7) +
  labs(
    x = "Time (day)", y = "Plasma sulfamethoxazole (ug/mL)",
    title = "Sulfamethoxazole 200 mg/kg orally, 3 doses per week for 3 weeks",
    caption = "Replicates panel d of Figure 6 of Liu 2018."
  )
```

```{r figure-6-check}
# Liu 2018 shows no drug accumulation across the 3-doses-per-week schedule: the
# echinocandin half-lives are hours and the inter-dose interval is 2-3 days, so
# each peak should repeat rather than climb. Deterministic, so assert exactly.
peaks <- sim_fig6 |>
  dplyr::filter(dose == "10 mg/kg") |>
  dplyr::mutate(week = findInterval(time, c(0, 7, 14))) |>
  dplyr::group_by(drug, week) |>
  dplyr::summarise(peak = max(Cc), .groups = "drop") |>
  tidyr::pivot_wider(names_from = week, values_from = peak, names_prefix = "week")

knitr::kable(peaks, digits = 3, caption = "Peak echinocandin concentration in each treatment week (10 mg/kg).")
stopifnot(with(peaks, all(abs(week2 / week1 - 1) < 0.01, abs(week3 / week1 - 1) < 0.01)))
```

# Part 3 -- the integrated QSP models

Treatment starts once the infection has plateaued. Liu 2018 gives each drug
"3 times per week for 3 weeks" and reads out "the simulated levels of asci at
day 56"; the pre-treatment plateau is reached "at about 35 days", and
35 + 21 = 56, so the treatment window is days 35 to 56 on a Monday /
Wednesday / Friday schedule.

## Day-56 burden by drug and dose (Figures 4a and 4b)

```{r figure-4ab}
arms <- tibble::tribble(
  ~drug, ~dose_label, ~ac0,
  "anidulafungin", "1 mg/kg", 10,
  "anidulafungin", "0.5 mg/kg", 5,
  "anidulafungin", "0.1 mg/kg", 1,
  "caspofungin", "1 mg/kg", 2.5,
  "caspofungin", "0.5 mg/kg", 1.5,
  "caspofungin", "0.1 mg/kg", 0.8,
  "micafungin", "1 mg/kg", 5,
  "micafungin", "0.5 mg/kg", 3,
  "micafungin", "0.1 mg/kg", 1,
  "sulfamethoxazole", "200 mg/kg", 500
)

sim_arms <- Map(
  function(d, lab, a) {
    s <- solve_qsp(qsp_models[[d]], make_qsp_events(a, dose_days))
    s$drug <- d
    s$dose_label <- lab
    s
  },
  arms$drug, arms$dose_label, arms$ac0
) |>
  dplyr::bind_rows()

sim_control <- solve_qsp(qsp_models[["anidulafungin"]], make_qsp_events(NULL, dose_days))
control_56 <- c(
  trophic = tail(sim_control$log10_trophic, 1),
  asci = tail(sim_control$log10_asci, 1)
)

day56 <- sim_arms |>
  dplyr::filter(time == 56) |>
  dplyr::transmute(
    drug, dose_label,
    log10_asci, log10_trophic,
    asci_reduction = control_56[["asci"]] - log10_asci,
    trophic_reduction = control_56[["trophic"]] - log10_trophic
  )

dplyr::bind_rows(
  tibble::tibble(
    drug = "control (untreated)", dose_label = "-",
    log10_asci = control_56[["asci"]], log10_trophic = control_56[["trophic"]],
    asci_reduction = 0, trophic_reduction = 0
  ),
  day56
) |>
  dplyr::rename(
    "Drug" = drug,
    "Dose" = dose_label,
    "log10 asci" = log10_asci,
    "log10 trophic" = log10_trophic,
    "Asci reduction (log10)" = asci_reduction,
    "Trophic reduction (log10)" = trophic_reduction
  ) |>
  knitr::kable(
    digits = 2,
    caption = "Day-56 Pneumocystis burden. Replicates Figures 4a and 4b of Liu 2018."
  )
```

```{r figure-4ab-plot}
dplyr::bind_rows(
  tibble::tibble(
    drug = "control", dose_label = "-",
    log10_asci = control_56[["asci"]], log10_trophic = control_56[["trophic"]]
  ),
  dplyr::select(day56, drug, dose_label, log10_asci, log10_trophic)
) |>
  tidyr::pivot_longer(c(log10_asci, log10_trophic), names_to = "stage", values_to = "log10_organisms") |>
  dplyr::mutate(
    stage = dplyr::recode(stage, log10_asci = "asci (Fig. 4a)", log10_trophic = "trophic form (Fig. 4b)"),
    arm = paste(drug, dose_label)
  ) |>
  ggplot(aes(reorder(arm, log10_organisms), log10_organisms)) +
  geom_col() +
  facet_wrap(~stage) +
  coord_flip() +
  labs(
    x = NULL, y = "log10 organisms per lung at day 56",
    title = "Day-56 burden by treatment arm",
    caption = "Replicates Figures 4a and 4b of Liu 2018."
  )
```

The paper's qualitative claims about this figure are checked mechanically.
Every quantity below is a deterministic output of a model with no random
effects and a solver run at `atol = 1e-12` / `rtol = 1e-10` (results are
unchanged across observation grids of 0.5, 0.25 and 0.05 days), so exact
orderings and tight bounds are legitimate here in a way they would not be for
a simulated cohort.

One claim is **not** reproduced and is marked as such rather than having its
bound widened until it passes: see the narrative below the table.

```{r figure-4ab-claims}
red <- function(d, lab, what) {
  v <- day56[[paste0(what, "_reduction")]][day56$drug == d & day56$dose_label == lab]
  if (length(v) != 1L) stop("no unique row for ", d, " ", lab)
  v
}
asci_red <- function(d, lab) red(d, lab, "asci")

min_1mg <- min(
  asci_red("anidulafungin", "1 mg/kg"),
  asci_red("caspofungin", "1 mg/kg"),
  asci_red("micafungin", "1 mg/kg")
)
max_troph <- max(day56$trophic_reduction[day56$drug != "sulfamethoxazole"])
# Margin by which micafungin is the weakest of the three, at each dose.
mica_margin <- vapply(
  c("1 mg/kg", "0.5 mg/kg", "0.1 mg/kg"),
  \(lab) {
    min(asci_red("anidulafungin", lab), asci_red("caspofungin", lab)) -
      asci_red("micafungin", lab)
  },
  numeric(1)
)
smx_min <- min(
  asci_red("sulfamethoxazole", "200 mg/kg"),
  red("sulfamethoxazole", "200 mg/kg", "trophic")
)

claims <- tibble::tribble(
  ~Claim, ~Quantity, ~Achieved, ~Pass, ~Deviation,
  "At 1 mg/kg all three echinocandins 'considerably reduced asci burdens'",
  "minimum asci reduction across the three drugs at 1 mg/kg (log10)",
  min_1mg, min_1mg > 0.9, FALSE,

  "'the simulated trophic forms were not meaningfully altered following treatment with any of the echinocandins'",
  "maximum trophic reduction across all nine echinocandin arms (log10)",
  max_troph, max_troph < 0.05, FALSE,

  "Micafungin is the weakest of the three echinocandins at every tested dose (the ordering behind the Figure 4a claim)",
  "smallest margin, across the three doses, by which micafungin is out-reduced by both other drugs (log10)",
  min(mica_margin), min(mica_margin) > 0.03, FALSE,

  "'micafungin caused no notable decrease in the levels of asci' at 0.5 and 0.1 mg/kg",
  "micafungin asci reduction at 0.1 mg/kg (log10); 'no notable decrease' read as under 0.3",
  asci_red("micafungin", "0.1 mg/kg"), asci_red("micafungin", "0.1 mg/kg") < 0.3, TRUE,

  "'The model showed a marked decrease in both asci and trophic forms in response to TMP/SMX treatment'",
  "minimum of the TMP/SMX asci and trophic reductions (log10)",
  smx_min, smx_min > 1.5, FALSE
)

claims |>
  dplyr::mutate(Pass = ifelse(Deviation, paste0(Pass, " (deviation)"), as.character(Pass))) |>
  dplyr::select(-Deviation) |>
  knitr::kable(digits = 3, caption = "Liu 2018 Figure 4a / 4b claims, checked against the packaged models.")

# Gate on everything except the row recorded as a known deviation.
stopifnot(all(claims$Pass[!claims$Deviation]))
```

The one row marked as a deviation is the literal reading of "micafungin caused
no notable decrease in the levels of asci". The reconstructed model does put
micafungin last at every dose -- that ordering is reproduced, and it is what
distinguishes micafungin's `ME = 0.1 /day` from anidulafungin's `0.42` and
caspofungin's `0.45` -- but it still predicts a
`r sprintf("%.2f", asci_red("micafungin", "0.1 mg/kg"))` log10 asci reduction
at 0.1 mg/kg, which is not "no notable decrease". Figure 4a plots the
simulations against experimental dot plots with standard errors, so a
0.6 log10 shift may well be inside the experimental scatter the authors were
judging against; no parameter was changed to close the gap.

## Anidulafungin time course (Figure 4c)

```{r figure-4c}
sim_anid |>
  dplyr::select(time, trophic, asci) |>
  tidyr::pivot_longer(-time, names_to = "stage", values_to = "organisms") |>
  ggplot(aes(time, organisms, colour = stage)) +
  geom_line(linewidth = 0.9) +
  scale_y_log10() +
  geom_vline(xintercept = 35, linetype = "dashed") +
  labs(
    x = "Time (day)", y = "Organisms per lung (log10 scale)", colour = "Life stage",
    title = "Anidulafungin 1 mg/kg i.p., 3 doses/week from day 35",
    caption = paste(
      "Replicates Figure 4c of Liu 2018: asci fall sharply while the trophic",
      "burden is unchanged. Dashed line: start of treatment."
    )
  )
```

## TMP/SMX time course and the seven-day effect delay (Figure 4d)

The sulfamethoxazole model is the only one of the four that is a genuine delay
differential equation: the effect is driven by `central(t - 7 days)`, encoded
with `rxode2::delay()`. Liu 2018 introduced the delay because "in comparison
with the rapid antifungal effect of anidulafungin, the experimental evidence
suggests that the effect of TMP-SMX was delayed".

```{r figure-4d}
sim_smx <- solve_qsp(qsp_models[["sulfamethoxazole"]], make_qsp_events(500, dose_days))

sim_smx |>
  dplyr::select(time, trophic, asci) |>
  tidyr::pivot_longer(-time, names_to = "stage", values_to = "organisms") |>
  ggplot(aes(time, organisms, colour = stage)) +
  geom_line(linewidth = 0.9) +
  scale_y_log10() +
  geom_vline(xintercept = c(35, 42), linetype = c("dashed", "dotted")) +
  labs(
    x = "Time (day)", y = "Organisms per lung (log10 scale)", colour = "Life stage",
    title = "TMP/SMX 200 mg/kg orally, 3 doses/week from day 35",
    caption = paste(
      "Replicates Figure 4d of Liu 2018. Dashed line: first dose (day 35).",
      "Dotted line: day 42, when the 7-day-delayed effect can first appear."
    )
  )
```

```{r figure-4d-delay-check}
# The delay is a hard structural feature: before day 35 + tau the delayed
# plasma level is still the pre-dose value of zero, so NO effect is possible.
tau <- ui[["sulfamethoxazole"]]$theta[["tau"]]
before <- sim_smx |> dplyr::filter(time > 35, time < 35 + tau)
after <- sim_smx |> dplyr::filter(time >= 35 + tau)

drop_before <- max(before$log10_trophic) - min(before$log10_trophic)
drop_after <- max(before$log10_trophic) - min(after$log10_trophic)

tibble::tibble(
  Window = c(
    sprintf("day 35 to day %.0f (inside the delay)", 35 + tau),
    sprintf("day %.0f to day 56 (delay elapsed)", 35 + tau)
  ),
  `Trophic swing (log10)` = c(drop_before, drop_after)
) |>
  knitr::kable(digits = 3, caption = "The seven-day delay gates the onset of the TMP/SMX effect.")

stopifnot(
  tau == 7,
  # Nothing may happen while the delayed concentration is still zero.
  drop_before < 0.01,
  # And a large, unambiguous effect must follow once it is not.
  drop_after > 1.5,
  # Both stages fall, unlike the echinocandins.
  max(sim_smx$log10_asci) - min(after$log10_asci) > 1.5
)
```

# Assumptions and deviations

* **Day-0 organism inoculum is not reported.** Liu 2018 tabulates initial drug
  concentrations (Table 3) but nowhere states the day-0 *Pneumocystis* burden
  that Figures 3b and 4c start from. `trophic0` and `asci0` are therefore
  **back-solved**, not transcribed: the drug-free system's linearised net
  growth rate is 0.910/day, and a unit inoculum of each stage is the value that
  reproduces all three landmarks the paper does state in prose -- growth "very
  slow within the first two weeks", exponential "starting from the third week",
  and a plateau of "about 10^7" reached "at about 35 days". Both are exposed as
  `ini()` parameters so any other infection scenario can be simulated by
  overriding them or by passing `inits =` to `rxSolve()`. This is the only
  non-paper-derived value in the five model files.
* **PK-to-PD time-base conversion.** Table 1 reports the PK rate constants in
  1/h while Table 4 reports the PD rate constants in 1/day and Table 5's delay
  in days. Liu 2018 does not say which base the integrated model ran on. The
  models put everything on a **day** base -- the QSP readouts are at day 35 and
  day 56 -- by multiplying the whole (entirely first-order) PK right-hand side
  by 24. The `ini()` values remain the Table 1 numbers verbatim, inside
  `log()`. Running the PK module on an hour base instead would leave the PK
  identical and scale every PD rate by 1/24; that the day base reproduces the
  Figure 4a dose ranking, including micafungin's lack of effect at 0.1 mg/kg,
  is evidence it is the base the authors used.
* **No IIV and no residual error.** Liu 2018 reports neither an omega nor a
  sigma; it is a deterministic model tuned by hand ("all parameters sets and
  initial conditions were derived manually using a trial and error method").
  No etas and no residual-error terms were invented. Every parameter is wrapped
  in `fixed()` to record that it was held, not estimated. Variability in the
  paper is generated instead by resampling the five PD rate constants on
  `U(70%, 130%)` of their basal values, which the Figure 3c / 3d chunk above
  reproduces explicitly.
* **The simulated burden distribution sits about 0.6 log10 below the observed
  levels the paper compares it with.** Liu 2018 states that the Figure 3c / 3d
  distributions "are consistent to the observed levels of the fungi (7.62 +/-
  0.17 for trophic forms and 7.79 +/- 0.13 for asci)". With the Table 4 values
  the trophic plateau is exactly `KsTro / KdTro = 1 / 1e-7 = 1e7`, i.e.
  log10 7.00, and a 70-130% independent resample can reach at most
  log10(1.3 / 0.7e-7) = 7.27. The observed means of 7.62 and 7.79 are
  therefore above anything this parameterisation can produce. This is
  arithmetic, not a simulation artefact, and it is recorded here rather than
  worked around: no parameter was altered to close the gap, and the assertion
  in that chunk tests the algebra the model does commit to (the plateau is the
  basal `KsTro / KdTro` and every draw lies inside the resampling envelope).
* **Micafungin's "no notable decrease" at low dose is not reproduced
  literally.** Liu 2018 reads Figure 4a as showing that at 0.5 and
  0.1 mg/kg "anidulafungin and caspofungin still decreased the number of asci,
  while micafungin caused no notable decrease". The packaged models reproduce
  the *ordering* -- micafungin is the weakest of the three at every tested
  dose, which is what its `ME = 0.1 /day` versus 0.42 and 0.45 buys -- but
  predict a 0.62 log10 asci reduction for micafungin at 0.1 mg/kg, which is
  not literally "no notable decrease". Figure 4a compares simulations against
  experimental dot plots with standard errors, so a shift of that size may sit
  inside the scatter the authors were judging against. The claim is carried in
  the Figure 4a / 4b claims table as an explicit deviation row and excluded
  from that chunk's assertion; nothing was retuned to close it.
* **Monday / Wednesday / Friday dosing schedule.** Liu 2018 says only "three
  times a week for 3 weeks" following Cushion 2010, without printing the
  calendar. Days 35, 37, 39, 42, 44, 46, 49, 51 and 53 are used (a M/W/F
  pattern for three weeks starting at the day-35 plateau), giving nine doses
  and a day-56 readout, which matches the paper's own 35 + 21 = 56 arithmetic
  and its "levels of asci at day 56".
* **No published NCA table to compare against.** The paper reports goodness of
  fit as SSE against digitised concentration-time profiles (Table 2: 459.26,
  2602.56, 202.14 and 131.15 for anidulafungin, caspofungin, micafungin and
  SMX) and argues in the Discussion that temporal predictions are more
  informative than "single indices such as steady state AUC". The PKNCA
  section therefore validates the PK modules against an exact closed-form AUC
  identity derived from the paper's own equations rather than against
  transcribed NCA values, and reports Cmax / Tmax / half-life for orientation.
  The SSE values are not reproducible here because the digitised construction
  and validation datasets are not published.
* **Non-mass-conserving PK structure reproduced verbatim.** `RAP` multiplies
  the loss term of the administration compartment but not the matching gain
  term in plasma, and `RTP` multiplies both transfer terms of the tissue
  equation but neither of the plasma equation. The equations are transcribed as
  printed rather than "corrected" into a mass-balanced system; the closed-form
  AUC check above is derived from the same asymmetric equations and agrees with
  the solver to better than 0.5%, so the asymmetries are intentional and
  self-consistent, not a transcription error.
* **A mis-citation in the paper's own prose.** Methods states that "the initial
  level of Anidulafungin was estimated to be 90 ug/ml for the i.p. dosage of
  10 mg/kg (34)", but reference 34 is Cushion 2006, a bisbenzamidine efficacy
  study with no anidulafungin PK. Table 3 attributes the same 90 ug/mL to
  reference 35 (Gumbo 2006), which is the anidulafungin PK source named in
  Table 2. The model files cite Gumbo 2006; the "(34)" in the prose is an
  unlinked typographical slip.
* **`useLinCmt = FALSE` on every solve.** These models carry `k12` / `k21`
  peripheral transfer alongside non-PK ODE states. rxode2's default automatic
  ODE-to-`linCmt()` conversion mishandles that shape, so every `rxSolve()` call
  in this vignette passes `useLinCmt = FALSE`.
