---
title: "Dioxin-like compound mixture PBPK (Liu 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dioxin-like compound mixture PBPK (Liu 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

```{r libraries, message = FALSE}
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Liu, Zacharewski, Conolly and Zhang (2022, *Toxics* 10(11):700,
[doi:10.3390/toxics10110700](https://doi.org/10.3390/toxics10110700)) present a
physiologically based pharmacokinetic (PBPK) **modeling framework for mixtures
of dioxin-like compounds (DLCs)** in humans. DLCs (polychlorinated
dibenzo-*p*-dioxins, dibenzofurans and biphenyls, with 2,3,7,8-TCDD as the
reference congener) are persistent lipophilic pollutants taken up chronically
through the diet. Their toxicokinetics are nonlinear: in the liver each congener
binds reversibly to the aryl hydrocarbon receptor (AHR) and to CYP1A2, and the
AHR-liganded congener induces CYP1A2, which both sequesters and metabolises the
congeners. In a mixture, congeners therefore compete for AHR and CYP1A2 and
**cross-induce** CYP1A2 - an interaction that cannot be captured by simulating
each congener alone.

The framework adapts the human TCDD model of Emond et al. Blood is a
quasi-steady-state algebraic compartment; fat, rest-of-body and liver are each
split into a blood subcompartment and a diffusion-limited tissue subcompartment.
Liver tissue carries the AHR / CYP1A2 binding and the CYP1A2 mRNA -> protein
induction cascade. Body weight and all tissue volume / flow fractions are
polynomial functions of age, so each run is a **lifetime (0-70 year) forward
simulation of constant daily dietary intake** - there are no dose records and
the model is deterministic.

The authors implemented the model in GNU Octave and published the full source
(parameters, ODEs and the chemical-specific parameter spreadsheet) at
<https://github.com/pulsatility/2022-DLC-Mixture-PBPK>, which the paper
(Section 2.6) designates as the model source. The parameter values and equations
used here are transcribed from that code (`DLC_human_cmd.m`, `DLC_human_ode.m`,
`DLC_human_parameters.xlsx`; paper Tables S1-S3).

Two instances of the framework are shipped:

* `Liu_2022_tcdd_pbpk` - the single-congener (TCDD) model, the framework's
  validation case (paper Figure 2, which reproduces the Emond model exactly).
* `Liu_2022_dioxin_mixture_pbpk` - a binary TCDD + 2,3,4,7,8-PeCDF mixture,
  demonstrating the cross-induction result (paper Figure 3A).

```{r load-models}
tcdd <- rxode2::rxode2(readModelDb("Liu_2022_tcdd_pbpk"))
mix <- rxode2::rxode2(readModelDb("Liu_2022_dioxin_mixture_pbpk"))
```

* Citation: `r readModelDb("Liu_2022_tcdd_pbpk")()$reference`

# Population

This is a mechanistic forward-simulation model, not a fit to a clinical cohort.
The physiology is the Emond human model (tissue volumes and blood flows as
polynomial functions of age; Wang et al. 1997), run for a female from birth to
70 years. Exposure is a constant daily dietary intake per congener, driven at
the EPA reference dose (RfD, 0.0007 ng/kg bw/day) and point-of-departure dose
(PoD, 0.02 ng/kg bw/day) lifetime scenarios. The paper's population Monte Carlo
analysis (its Figure 5) samples total AHR, CYP1A2 transcription / basal levels,
CYP1A2 Emax, the body-weight polynomial and the dietary intake of 11 congeners;
those distributions are summarised in the model `population` metadata but are not
encoded as random effects here (the model is deterministic).

```{r population}
readModelDb("Liu_2022_tcdd_pbpk")()$population[c("species", "age_range", "dose_range")]
```

# Source trace

Every `ini()` parameter carries an in-file comment pointing to its origin in the
authors' Octave code / parameter spreadsheet. Key structural choices:

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Blood (quasi-steady state, algebraic) | - | `DLC_human_ode.m` `CB` |
| Diffusion-limited fat / RB / liver tissue | - | `DLC_human_ode.m` ODEs 5-10 |
| AHR binding `Kd,AHR` (TCDD) | 0.1 nM | `DLC_human_parameters.xlsx` `KDAHR` |
| `kb,AHR` | 6 /h | paper Section 2.3 ("we set it to 6/h") |
| CYP1A2 binding `Kd,1A2` | 40 nM | `xlsx` `KD1A2` |
| CYP1A2 induction Emax / EC50 / Hill | 9300 / 130 nM / 0.6 | `DLC_human_cmd.m`, `DLC_human_ode.m` |
| CYP1A2 basal (initial) | 1600 nM | `DLC_human_cmd.m` `CYP1A2_1BASAL` |
| Hepatic elimination `kelim` (TCDD) | 0.0011 /h | `xlsx` `kelim` |
| Partition coefficients `PL`/`PF`/`PRB` (TCDD) | 6 / 100 / 1.5 | `xlsx` |
| Growth / volume polynomials (female) | - | `DLC_human_ode.m` |

# Validation against the authors' reference implementation

The strongest available check is reproduction of the authors' own Octave model.
Running `DLC_human_ode.m` (verbatim) for the single-TCDD and the binary
TCDD + PeCDF cases yields the tissue concentrations below at the eight ages the
paper scatter-plots in Figure 2B. The packaged rxode2 models must reproduce
these to integrator tolerance (the two implementations share identical
parameters, so any difference is pure numerical error).

```{r reference-data}
# Reference output of the authors' Octave code (DLC_human_ode.m), female.
# Columns: year, CA, CF, CRB, CL (total liver), CYP1A2 (free, nM).
ref_tcdd_high <- as.data.frame(do.call(rbind, list(
  c(1, 0.0004752823385, 0.04585359264, 0.0007124136162, 0.2261675182, 18730.81163),
  c(10, 0.001683453231, 0.1669597553, 0.002524947414, 1.609063533, 37940.59732),
  c(20, 0.002019160798, 0.2002846614, 0.003028699096, 2.136984426, 42044.22034),
  c(30, 0.002282740832, 0.2273015476, 0.003424082144, 2.58723345, 45055.86035),
  c(40, 0.002532412987, 0.253053741, 0.003798593922, 3.04158104, 47771.19626),
  c(50, 0.002693638141, 0.2696631966, 0.004040442755, 3.34889642, 49462.93963),
  c(60, 0.002725673327, 0.2728344271, 0.004088514447, 3.411510536, 49795.40668),
  c(70, 0.002615174887, 0.2612254096, 0.003922785398, 3.198696246, 48649.61923)
)))
names(ref_tcdd_high) <- c("year", "CA", "CF", "CRB", "CL", "CYP")

ref_tcdd_low <- as.data.frame(do.call(rbind, list(
  c(1, 2.859622848e-05, 0.002752792668, 4.286141748e-05, 0.003692457412, 4789.488696),
  c(10, 0.0001375501738, 0.01363323401, 0.0002063040658, 0.03493473979, 9784.584703),
  c(20, 0.00017018765, 0.01687623487, 0.0002552772614, 0.04796207462, 10898.09831),
  c(30, 0.0002051219412, 0.02039887323, 0.0003076780594, 0.063434271, 11996.81454),
  c(40, 0.0002531608954, 0.02525558411, 0.0003797354479, 0.08710258389, 13390.92554),
  c(50, 0.0003004094215, 0.03003377966, 0.0004506089893, 0.1128851484, 14660.81984),
  c(60, 0.0003275725556, 0.03277036245, 0.0004913575589, 0.1287697037, 15354.46011),
  c(70, 0.0003161010399, 0.0315891858, 0.0004741557633, 0.1219833922, 15065.2614)
)))
names(ref_tcdd_low) <- c("year", "CA", "CF", "CRB", "CL", "CYP")

# Binary TCDD + PeCDF, each at PoD (0.02 ng/kg/day).
# Columns: year, TCDD[CA CF CRB CL], PeCDF[CA CF CRB CL], CYP1A2.
ref_bin <- as.data.frame(do.call(rbind, list(
  c(1, 0.00042756585, 0.0412646, 0.00064089545, 0.24297473, 0.00034990559, 0.045288913, 0.0011063204, 0.20597163, 22460.876),
  c(10, 0.001435269, 0.14236774, 0.002152711, 1.6745916, 0.0014517792, 0.19510039, 0.0045962913, 1.7244181, 46422.124),
  c(20, 0.00169754, 0.16840313, 0.0025462772, 2.2252945, 0.0018897718, 0.2540512, 0.00598404, 2.5177863, 52195.479),
  c(30, 0.0018866034, 0.18791254, 0.0028298856, 2.668139, 0.0022526962, 0.30434759, 0.0071333451, 3.2345621, 56347.02),
  c(40, 0.0020432841, 0.20425191, 0.0030649115, 3.0838711, 0.0026468183, 0.35926375, 0.0083813915, 4.0523659, 60162.82),
  c(50, 0.0021213545, 0.21243485, 0.0031820259, 3.3405127, 0.0029818266, 0.4058321, 0.0094422953, 4.7607909, 62788.987),
  c(60, 0.0021115052, 0.21138364, 0.0031672638, 3.3716316, 0.003151567, 0.42905952, 0.0099799323, 5.1017177, 63672.551),
  c(70, 0.0020252708, 0.20227637, 0.0030379217, 3.1712343, 0.0030607917, 0.41579757, 0.0096926411, 4.8601557, 62426.629)
)))
names(ref_bin) <- c("year", "CA1", "CF1", "CRB1", "CL1", "CA2", "CF2", "CRB2", "CL2", "CYP")
```

```{r solve-helpers}
# Forward lifetime solve on a half-yearly grid (hours). Tight tolerances so the
# comparison against the reference implementation reflects the model, not the
# integrator.
tgrid <- seq(0, 24 * 365 * 70, by = 24 * 365 / 2)
ev <- rxode2::et(tgrid)

solve_tcdd <- function(mod, mstot) {
  s <- rxode2::rxSolve(mod, ev, params = c(MSTOT = mstot),
    atol = 1e-12, rtol = 1e-10, method = "lsoda")
  s$year <- s$time / 24 / 365
  s
}

s_high <- solve_tcdd(tcdd, 0.02)
s_low <- solve_tcdd(tcdd, 7e-4)
```

```{r validate-tcdd}
# Match the rxode2 solve to the reference ages and compute the relative error.
rel_err_at_ages <- function(sim, ref, cols) {
  idx <- vapply(ref$year, function(y) which.min(abs(sim$year - y)), integer(1))
  errs <- vapply(seq_along(cols), function(k) {
    max(abs(sim[[cols[[k]]]][idx] - ref[[names(cols)[k]]]) / ref[[names(cols)[k]]])
  }, numeric(1))
  stats::setNames(errs, names(cols))
}

cols_tcdd <- c(CA = "Cc", CF = "Cfat", CRB = "Crest", CL = "Cliver_total", CYP = "Ccyp1a2")
err_high <- rel_err_at_ages(s_high, ref_tcdd_high, cols_tcdd)
err_low <- rel_err_at_ages(s_low, ref_tcdd_low, cols_tcdd)
print(round(rbind(PoD = err_high, RfD = err_low), 9))

stopifnot(
  "TCDD PoD reproduces the Octave reference" = max(err_high) < 1e-4,
  "TCDD RfD reproduces the Octave reference" = max(err_low) < 1e-4
)
```

```{r validate-binary}
sb <- rxode2::rxSolve(mix, ev,
  params = c(MSTOT_tcdd = 0.02, MSTOT_pecdf = 0.02),
  atol = 1e-12, rtol = 1e-10, method = "lsoda")
sb$year <- sb$time / 24 / 365

cols_t <- c(CA1 = "Cc_tcdd", CF1 = "Cfat_tcdd", CRB1 = "Crest_tcdd", CL1 = "Cliver_total_tcdd")
cols_p <- c(CA2 = "Cc_pecdf", CF2 = "Cfat_pecdf", CRB2 = "Crest_pecdf", CL2 = "Cliver_total_pecdf")
err_bt <- rel_err_at_ages(sb, ref_bin, cols_t)
err_bp <- rel_err_at_ages(sb, ref_bin, cols_p)
err_bc <- rel_err_at_ages(sb, ref_bin, c(CYP = "Ccyp1a2"))
print(round(rbind(TCDD = err_bt, PeCDF = err_bp), 9))

stopifnot(
  "binary TCDD reproduces the Octave reference" = max(err_bt) < 1e-4,
  "binary PeCDF reproduces the Octave reference" = max(err_bp) < 1e-4,
  "binary CYP1A2 reproduces the Octave reference" = max(err_bc) < 1e-4
)
```

# Lifetime accumulation and CYP1A2 induction (Figure 2)

For single TCDD exposure, tissue burdens accumulate over the lifetime and
CYP1A2 is induced far above its basal level (1600 nM). The PoD scenario drives
~30-fold CYP1A2 induction; the RfD scenario, ~9-fold.

```{r fig2, fig.cap = "Lifetime TCDD tissue concentrations and liver CYP1A2 induction (reproduces Liu 2022 Figure 2A). Points are the authors' Octave reference output."}
traj <- bind_rows(
  transform(s_high[, c("year", "Cc", "Cfat", "Crest", "Cliver_total", "Ccyp1a2")], dose = "PoD (0.02)"),
  transform(s_low[, c("year", "Cc", "Cfat", "Crest", "Cliver_total", "Ccyp1a2")], dose = "RfD (0.0007)")
) |>
  pivot_longer(c(Cc, Cfat, Crest, Cliver_total, Ccyp1a2), names_to = "output", values_to = "conc")

refpts <- bind_rows(
  transform(ref_tcdd_high, dose = "PoD (0.02)"),
  transform(ref_tcdd_low, dose = "RfD (0.0007)")
) |>
  rename(Cc = CA, Cfat = CF, Crest = CRB, Cliver_total = CL, Ccyp1a2 = CYP) |>
  pivot_longer(c(Cc, Cfat, Crest, Cliver_total, Ccyp1a2), names_to = "output", values_to = "conc")

ggplot(traj, aes(year, conc, colour = dose)) +
  geom_line() +
  geom_point(data = refpts, size = 1) +
  facet_wrap(~output, scales = "free_y") +
  labs(x = "Age (years)", y = "Concentration (nM)", colour = "Dietary intake") +
  theme_bw()
```

# Mixture cross-induction lowers extrahepatic burden (Figure 3A)

The paper's central result: when two congeners are co-administered, shared
CYP1A2 is more strongly induced than under either single exposure, which
increases hepatic sequestration and metabolism and so **lowers each congener's
extrahepatic (blood, fat, rest-of-body) burden** relative to single exposure.
The liver burden is the exception - it can rise. Below, TCDD is simulated alone
and in the binary mixture with 2,3,4,7,8-PeCDF.

```{r fig3}
s_tcdd_single <- s_high # TCDD alone at PoD

# lifetime-averaged mixture:single ratio (excluding t = 0), as in the paper.
ratio <- function(mix_col, single_col) {
  m <- sb[[mix_col]][-1]
  s <- s_tcdd_single[[single_col]][-1]
  mean(m / s)
}
ratios_tcdd <- c(
  CA = ratio("Cc_tcdd", "Cc"),
  CF = ratio("Cfat_tcdd", "Cfat"),
  CRB = ratio("Crest_tcdd", "Crest"),
  CL = ratio("Cliver_total_tcdd", "Cliver_total")
)
print(round(ratios_tcdd, 3))
```

```{r fig3-assert}
# The paper (Section 3.2, Figure 3): CA, CF and CRB ratios are below unity
# (extrahepatic reduction); CYP1A2 is more induced in the mixture than alone.
stopifnot(
  "extrahepatic TCDD burden is reduced in the mixture" =
    all(ratios_tcdd[c("CA", "CF", "CRB")] < 1),
  "mixture induces CYP1A2 above single exposure" =
    tail(sb$Ccyp1a2, 1) > tail(s_tcdd_single$Ccyp1a2, 1)
)
cat(sprintf(
  "lifetime-end CYP1A2: single %.0f nM, binary %.0f nM\n",
  tail(s_tcdd_single$Ccyp1a2, 1), tail(sb$Ccyp1a2, 1)
))
```

```{r fig3-plot, fig.cap = "TCDD blood (CA) and total liver (CL) concentrations, single vs binary mixture with 2,3,4,7,8-PeCDF (reproduces the pattern of Liu 2022 Figure 3A): blood burden falls in the mixture while liver burden is comparable."}
cmp <- bind_rows(
  data.frame(year = s_tcdd_single$year, CA = s_tcdd_single$Cc,
    CL = s_tcdd_single$Cliver_total, exposure = "TCDD single"),
  data.frame(year = sb$year, CA = sb$Cc_tcdd,
    CL = sb$Cliver_total_tcdd, exposure = "TCDD in mixture")
) |>
  pivot_longer(c(CA, CL), names_to = "tissue", values_to = "conc")

ggplot(cmp, aes(year, conc, colour = exposure)) +
  geom_line() +
  facet_wrap(~tissue, scales = "free_y") +
  labs(x = "Age (years)", y = "TCDD concentration (nM)", colour = NULL) +
  theme_bw()
```

# Full 11-congener parameter set

The framework scales to any number of congeners. The complete chemical-specific
parameter table (`DLC_human_parameters.xlsx`, paper Table S3) is reproduced
below so the shipped models can be re-parameterised for any of the 11 congeners
the paper studied. Columns are the liver / fat / rest-of-body partition
coefficients, the AHR dissociation constant, the hepatic elimination rate
constant, the GI absorption rate constant, molecular weight, intake toxic
equivalency factor (TEF) and systemic TEF.

```{r congener-table}
congeners <- data.frame(
  congener = c("2,3,4,7,8-PeCDF", "1,2,3,7,8-PeCDF", "2,3,7,8-TCDF", "1,2,3,4,7,8-HxCDF",
    "2,3,7,8-TCDD", "1,2,3,6,7,8-HxCDF", "1,2,3,4,6,7,8-HpCDD", "1,2,3,4,6,7,8-HpCDF",
    "1,2,3,7,8-PeCDD", "1,2,3,4,6,7,8,9-OCDD", "1,2,3,4,6,7,8,9-OCDF"),
  PL = c(27.551, 11.020, 11.020, 15.306, 6.000, 27.551, 20.816, 15.918, 10.408, 34.286, 9.184),
  PF = c(136.03, 52.63, 22.27, 30.36, 100.00, 52.63, 57.89, 56.28, 174.90, 22.27, 58.30),
  PRB = c(3.167, 0.833, 0.392, 0.917, 1.500, 2.083, 3.250, 0.608, 1.583, 2.167, 0.458),
  KDAHR = c(0.149, 0.769, 0.417, 2.326, 0.100, 2.703, 2.778, 0.588, 0.333, 285.714, 62.500),
  kelim = c(6.89e-4, 7.58e-3, 1.705e-2, 1.196e-3, 1.10e-3, 1.10e-3, 1.049e-3, 2.623e-3, 7.93e-4, 1.218e-3, 3.41e-2),
  KABS = c(0.0606, 0.0612, 0.0600, 0.0600, 0.0600, 0.0600, 0.0532, 0.0538, 0.0612, 0.0470, 0.0588),
  MW = c(340.4, 340.4, 306.0, 374.9, 322.0, 374.9, 425.3, 409.3, 356.4, 459.8, 443.8),
  TEF = c(0.3, 0.03, 0.1, 0.1, 1, 0.1, 0.01, 0.01, 1, 3e-4, 3e-4),
  Systemic_TEF = c(1, 0.1, 0.1, 1, 1, 0.1, 0.1, 0.01, 1, 3e-4, 1e-3)
)
knitr::kable(congeners, digits = 4)
```

# Assumptions and deviations

* **Parameter source.** The paper's numeric parameters live in supplements
  (Tables S1-S4) and in the authors' published Octave repository. The models
  here are transcribed from that repository
  (<https://github.com/pulsatility/2022-DLC-Mixture-PBPK>), which the paper
  designates as the model source, and were validated by reproducing the Octave
  output to integrator tolerance (section above).
* **Deterministic.** The model has no IIV and no residual error; the authors'
  population variability (Figure 5) comes from Monte Carlo sampling of AHR,
  CYP1A2 and body-weight parameters, not from estimated random effects. No error
  model is encoded.
* **Sex.** The results the paper presents are for women, and the shipped growth
  / fat-fraction polynomials are the female curves (`g.male = 0` in the Octave
  code). The deposit also contains male curves; they are not encoded here.
* **CYP1A2 intermediates.** The authors replaced the Emond model's two CYP1A2
  intermediate species with an explicit mRNA step, doubling the protein
  half-life to preserve the induction delay (paper Section 2.3). That mRNA
  formulation is what is encoded.
* **Fixed congener count.** rxode2 has no variable-length ODE vectors, so the
  single-congener and binary models fix the congener count at one and two. The
  full 11-congener parameter set is tabulated above for re-parameterisation.
* **Exposure encoding.** Daily dietary intake (ng/kg bw/day) enters as a
  continuous zero-order input into the GI compartment computed from the
  time-varying body weight; there are no dose records.
