---
title: "Selpercatinib (Liu 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Selpercatinib (Liu 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Liu_2025_selpercatinib"))$reference`
* Description: `r rxode2::rxode(readModelDb("Liu_2025_selpercatinib"))$description`
* Article: <https://doi.org/10.1002/psp4.70042>
* Supplement (Tables S1-S3, Figures S1-S7): <https://doi.org/10.1002/psp4.70042>
  (Supporting Information, `PSP4-14-1848-s001.docx`)

Selpercatinib is a highly selective RET kinase inhibitor. Liu 2025 updated a
previously developed adult population PK model with pediatric data in order to
choose a pediatric dose regimen by matching exposure to that of adults treated
at the approved 160 mg twice-daily dose. The resulting regimen (Table 3 of the
paper) was granted FDA accelerated approval for patients 2 years and older on
29 May 2024.

# Population

The analysis pooled two ongoing open-label phase 1/2 studies with a data cut-off
of 13 January 2023 (Liu 2025 Methods "Study Design", Table 1):

* **LIBRETTO-001** (NCT03157128), patients aged 12 years and older with advanced
  or metastatic solid tumors: 803 patients contributing 7723 concentrations.
  Doses ranged from 20 mg once daily to 240 mg twice daily during phase 1 dose
  escalation; phase 2 used 160 mg twice daily.
* **LIBRETTO-121** (NCT03899792), patients enrolled from 6 months to 21 years of
  age with an activating RET alteration and an advanced solid or primary CNS
  tumor: 27 patients contributing 301 concentrations, dosed on body surface area
  (92 mg/m^2 twice daily, capped at 160 mg twice daily).

Overall 830 patients contributed 8024 concentrations. Median age was 58.0 years
(range 2.0-92.0), median weight 67.0 kg (range 9.6-179) and median BSA 1.76 m^2
(range 0.446-2.79); 403/830 (48.6%) were female. Only 6 patients (0.7%) were
under 12 years of age and they contributed just 59 concentrations, which the
Discussion identifies as the analysis's principal limitation. A further 133
concentrations were below the 1 ng/mL assay limit of quantification and were
excluded.

Liu 2025 Table 1 does **not** report the race distribution, even though Asian
race is a covariate in the final model; see Assumptions below.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Liu_2025_selpercatinib")()$population`).

# Source trace

Every `ini()` entry carries an in-file comment naming its origin in
`inst/modeldb/specificDrugs/Liu_2025_selpercatinib.R`. They are collected here
for review. All final estimates come from Liu 2025 Table 2, "NONMEM parameter
estimates for the final model"; the covariate functions are printed in the
"Functions and Parameters Used to Test Covariates" block of that same table.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` = log CL/F | log(6.04) L/h | Table 2 `CL (L/h)` = 6.04 (95% CI 5.79/6.29) |
| `lvc` = log Vc/F | log(99.6) L | Table 2 `Vc (L)` = 99.6 (86.3/113) |
| `lvp` = log Vp/F | log(91.3) L | Table 2 `Vp (L)` = 91.3 (82.9/99.7) |
| `lq` = log Q/F | log(29.6) L/h | Table 2 `Q (L/h)` = 29.6 (24.2/35.0) |
| `lka` | log(1.47) 1/h | Table 2 `ka (1/h)` = 1.47 (1.25/1.69) |
| `ld1` = log zero-order duration | log(1.09) h | Table 2 `Dur (h)` = 1.09 (1.05/1.13) |
| `lfdepot` = log F1 | `fixed(log(1))` | Table 2 `F1 (fraction) 1.00 Fixed`; Figure 1 caption "F, bioavailability (fixed to 100%)" |
| `e_wt_cl` | `fixed(0.75)` | Methods "Population PK Analysis"; Table 2 covariate functions for CL and Q |
| `e_wt_vc` | `fixed(1)` | Methods "Population PK Analysis"; Table 2 covariate functions for Vc and Vp |
| `e_dose_cl` | -0.00321 /mg | Table 2 `Dose effect on CL (%/mg)` = -0.321 (-0.439/-0.203), divided by 100 |
| `e_race_asian_fdepot` | 0.183 | Table 2 `Asian race effect on F1 (%)` = 18.3 (10.9/25.8), divided by 100 |
| `etalcl` | 0.2136135 | Table 2 IIV on CL = 48.8%, as `log(CV^2 + 1)` |
| `etalvc` | 0.3625026 | Table 2 IIV on Vc = 66.1%, as `log(CV^2 + 1)` |
| `etalka` | 0.3396785 | Table 2 IIV on ka = 63.6%, as `log(CV^2 + 1)` |
| `etald1` | 0.2744783 | Table 2 IIV on Dur = 56.2%, as `log(CV^2 + 1)` |
| `addSd` | 61.5 ng/mL | Table 2 `Additive RUV (mg/L)` = 61.5 (37.4/85.6); unit tag read as ng/mL, see Errata |
| `propSd` | 0.253 | Table 2 `Proportional RUV (fraction)` = 0.253 (0.231/0.275) |
| Sequential zero- then first-order absorption; `dur(depot)`, then `ka` into `central` | n/a | Figure 1 schematic; Abstract; Results "PK Analyses" |
| Two-compartment disposition, first-order elimination | n/a | Figure 1 schematic; Methods "Population PK Analysis" |
| `CL = theta_CL * (1 + theta_dose * (Dose - 160 mg)) * (WT/70)^0.75` | n/a | Table 2 covariate-function block; Table S2 footnote (b) fixes the 160 mg centring |
| `F1 = theta_F1 * (1 + theta_Asian)` | n/a | Table 2 covariate-function block; Table S2 footnote (a) defines Asian vs non-Asian |
| `Vc = theta_Vc * (WT/70)^1.0`, `Vp = theta_Vp * (WT/70)^1.0`, `Q = theta_Q * (WT/70)^0.75` | n/a | Table 2 covariate-function block |
| Recommended pediatric regimen simulated below | n/a | Table 3 "Recommended selpercatinib doses for patients aged 2-17 years" |
| Adult reference exposure targets | Cmax ~3 mg/L, AUC0-24 ~53 mg*h/L | Figures 3 and 4, orange dashed "Adult reference level" line (read off the figure) |

# Closed-form gate on the typical-value model

Before any cohort is simulated, the structural model is checked against the
identity it must satisfy exactly. At steady state on a regimen of dose `D` every
`tau` hours, the area under the curve over one dosing interval is `F * D / CL`,
independent of absorption, distribution and the number of compartments. This is
a pure numerical check on the solved ODEs -- both sides use the same parameters,
so a tight bound is the right assertion here (unlike the cohort assertions
further down, which must be robust to which subjects are drawn).

```{r closed-form}
mod <- readModelDb("Liu_2025_selpercatinib")

# Typical 70 kg non-Asian adult on the approved 160 mg BID regimen, run out to
# 30 days so that the ~23 h terminal half-life is fully accumulated.
ev_typ <- rxode2::et(amt = 160, ii = 12, until = 24 * 30, rate = -2, cmt = "depot") |>
  rxode2::et(seq(24 * 29, 24 * 30, by = 0.02), cmt = "central") |>
  as.data.frame()
ev_typ$WT <- 70
ev_typ$DOSE <- 160
ev_typ$RACE_ASIAN <- 0

sim_typ <- rxode2::rxSolve(rxode2::zeroRe(mod), ev_typ, returnType = "data.frame")

# One dosing interval at the end of the run.
tau_win <- subset(sim_typ, time >= 24 * 29 + 12 & time <= 24 * 29 + 24)
auc_tau <- sum(diff(tau_win$time) *
                 (head(tau_win$Cc, -1) + tail(tau_win$Cc, -1)) / 2)
auc_closed <- 1 * 160 / 6.04 * 1000  # F * D / CL, in ng*h/mL

c(auc_solved = auc_tau, auc_closed_form = auc_closed,
  ratio = auc_tau / auc_closed)
```

```{r closed-form-gate}
# The two sides share the drawn parameters, so the only difference is
# trapezoidal and accumulation error -- a tight bound is correct here.
stopifnot(abs(auc_tau / auc_closed - 1) < 0.005)
```

The covariate functions are likewise checked as exact algebra rather than
eyeballed, using the per-subject parameters rxode2 returns alongside the
concentrations.

```{r covariate-regression}
grid <- expand.grid(WT = c(10, 30, 70, 120), DOSE = c(40, 120, 160, 240),
                    RACE_ASIAN = c(0, 1))
grid$id <- seq_len(nrow(grid))

ev_grid <- lapply(seq_len(nrow(grid)), function(i) {
  d <- rxode2::et(amt = grid$DOSE[i], ii = 12, until = 24, rate = -2, cmt = "depot") |>
    rxode2::et(c(0, 6, 12), cmt = "central") |>
    as.data.frame()
  d$id <- grid$id[i]
  d$WT <- grid$WT[i]
  d$DOSE <- grid$DOSE[i]
  d$RACE_ASIAN <- grid$RACE_ASIAN[i]
  d
}) |>
  dplyr::bind_rows()

sim_grid <- rxode2::rxSolve(rxode2::zeroRe(mod), ev_grid,
                            returnType = "data.frame") |>
  dplyr::distinct(id, .keep_all = TRUE) |>
  dplyr::left_join(grid, by = c("id", "WT", "DOSE", "RACE_ASIAN"))

chk <- sim_grid |>
  dplyr::mutate(
    cl_expected = 6.04 * (1 - 0.00321 * (DOSE - 160)) * (WT / 70)^0.75,
    vc_expected = 99.6 * (WT / 70),
    vp_expected = 91.3 * (WT / 70),
    q_expected  = 29.6 * (WT / 70)^0.75,
    f_expected  = 1 + 0.183 * RACE_ASIAN
  )

# Every covariate function must reproduce to machine precision.
stopifnot(
  max(abs(chk$cl / chk$cl_expected - 1)) < 1e-8,
  max(abs(chk$vc / chk$vc_expected - 1)) < 1e-8,
  max(abs(chk$vp / chk$vp_expected - 1)) < 1e-8,
  max(abs(chk$q  / chk$q_expected  - 1)) < 1e-8,
  max(abs(chk$fdepot / chk$f_expected - 1)) < 1e-8
)

# The dose effect reproduces the paper's direction and magnitude: apparent
# clearance falls as the dose rises, so exposure is more than dose
# proportional over the 20-240 mg range studied.
chk |>
  dplyr::filter(WT == 70, RACE_ASIAN == 0) |>
  dplyr::transmute(
    "Dose (mg)"          = DOSE,
    "CL/F (L/h)"         = round(cl, 3),
    "Multiplier vs 160 mg" = round(cl / 6.04, 4)
  ) |>
  knitr::kable(caption = "Table 2 dose effect on CL/F, -0.321% per mg, centred on 160 mg.")
```

# Virtual cohort

Original observed data are not publicly available. Liu 2025 drew its simulation
covariates from the NHANES DXA database, which is not reproduced here; instead
three arms are built whose covariate distributions match the published trial
demographics and whose dose assignment follows the paper's own rules.

```{r cohort}
# `set.seed()` seeds R's RNG, not rxode2's simulation streams (which are
# partitioned per solver thread), so this cohort is reproducible on one machine
# and different on a machine with a different thread count. Every assertion
# below is therefore written on medians and robust quantiles, never on the
# extremes of the drawn cohort.
set.seed(20250425)

N_PER_ARM <- 200L

# Growth reference used to turn an age into a weight and a height so that
# Mosteller BSA (and hence the Table 3 dose band) can be assigned. The weight
# curve is rescaled so that it passes through the LIBRETTO-121 median pair
# (14.0 years, 48.5 kg) from Liu 2025 Table 1; the height curve is a paediatric
# 50th-percentile reference. See Assumptions.
ref_age <- c(2, 3, 4, 5, 6, 8, 10, 12, 14, 16, 17)
ref_wt  <- c(12.5, 14.3, 16.3, 18.5, 20.8, 26.0, 32.5, 40.5, 50.0, 58.0, 61.0)
ref_ht  <- c(87, 96, 103, 110, 116, 128, 139, 150, 162, 170, 172)

wt_for_age <- function(age) exp(approx(log(ref_age), log(ref_wt), log(age), rule = 2)$y)
ht_for_age <- function(age) exp(approx(log(ref_age), log(ref_ht), log(age), rule = 2)$y)
wt_scale <- 48.5 / wt_for_age(14)

mosteller <- function(wt, ht) sqrt(wt * ht / 3600)

# Table 3 of Liu 2025: the recommended regimen for patients aged 2-17 years.
# 12 years and older is dosed on weight; 2 to under 12 years on BSA, with the
# 92 mg/m^2 target rounded to the available 40 and 80 mg capsules.
recommended_dose <- function(age, wt, bsa) {
  dplyr::case_when(
    age >= 12 & wt <  50 ~ 120,
    age >= 12 & wt >= 50 ~ 160,
    bsa >= 1.53          ~ 160,
    bsa >= 1.09          ~ 120,
    bsa >= 0.66          ~  80,
    TRUE                 ~  40   # BSA 0.33-0.65 m^2, given three times a day
  )
}
recommended_ii <- function(dose_mg) ifelse(dose_mg == 40, 8, 12)

# The comparator the paper rejected: adapted flat (adult weight-based) dosing
# applied to children as well.
flat_dose <- function(wt) ifelse(wt < 50, 120, 160)

make_peds <- function(n, regimen, id_offset = 0L) {
  age <- runif(n, 2, 17.99)
  wt  <- wt_for_age(age) * wt_scale * rlnorm(n, 0, 0.15)
  ht  <- ht_for_age(age) * rlnorm(n, 0, 0.04)
  bsa <- mosteller(wt, ht)
  amt <- if (regimen == "Recommended (Table 3)") {
    recommended_dose(age, wt, bsa)
  } else {
    flat_dose(wt)
  }
  tibble::tibble(
    id = id_offset + seq_len(n), AGE = age, WT = wt, BSA = bsa,
    DOSE = amt, ii = recommended_ii(amt), regimen = regimen,
    RACE_ASIAN = 0L
  )
}

# Adult reference arm: LIBRETTO-001 weight marginal (median 67.3 kg, SD 19.6,
# Liu 2025 Table 1) on the approved adult 160 mg BID regimen. Adults enter here
# only as the exposure target the pediatric regimen is matched to.
make_adults <- function(n, id_offset = 0L) {
  wt <- pmin(pmax(rlnorm(n, log(67.3), 0.27), 26.8), 179)
  tibble::tibble(
    id = id_offset + seq_len(n), AGE = NA_real_, WT = wt, BSA = NA_real_,
    DOSE = 160, ii = 12, regimen = "Adults 160 mg BID", RACE_ASIAN = 0L
  )
}

subjects <- dplyr::bind_rows(
  make_adults(N_PER_ARM, id_offset = 0L),
  make_peds(N_PER_ARM, "Recommended (Table 3)", id_offset = 1000L),
  make_peds(N_PER_ARM, "Adapted flat dosing",   id_offset = 2000L)
)

# Dose through the end of Cycle 1 Day 8 (day 1 spans 0-24 h, so day 8 spans
# 168-192 h), with a coarse grid through the accumulation phase and a fine grid
# across the C1D8 window that Figures 3 and 4 report.
build_records <- function(s) {
  dose_times <- seq(0, 192 - s$ii, by = s$ii)
  obs_times <- sort(unique(c(seq(0, 168, by = 4), seq(168, 192, by = 0.25))))
  dplyr::bind_rows(
    tibble::tibble(time = dose_times, evid = 1L, amt = s$DOSE,
                   rate = -2, cmt = "depot"),
    tibble::tibble(time = obs_times, evid = 0L, amt = NA_real_,
                   rate = NA_real_, cmt = "central")
  ) |>
    dplyr::mutate(id = s$id, AGE = s$AGE, WT = s$WT, BSA = s$BSA,
                  DOSE = s$DOSE, RACE_ASIAN = s$RACE_ASIAN,
                  regimen = s$regimen) |>
    dplyr::arrange(time, dplyr::desc(evid))
}

events <- subjects |>
  split(~ id) |>
  lapply(build_records) |>
  dplyr::bind_rows()

stopifnot(
  nrow(subjects) == 3L * N_PER_ARM,
  !anyDuplicated(subjects$id),
  all(events$DOSE %in% c(40, 80, 120, 160))
)

# Achieved cohort marginals against the published demographics.
subjects |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(
    n = dplyr::n(),
    "Median age (y)"    = round(median(AGE), 1),
    "Median weight (kg)" = round(median(WT), 1),
    "Median BSA (m2)"   = round(median(BSA), 2),
    "Median dose (mg)"  = median(DOSE),
    .groups = "drop"
  ) |>
  dplyr::rename(Regimen = regimen, N = n) |>
  knitr::kable(caption = "Simulated cohorts. The pediatric arms share their covariates and differ only in dose assignment.")
```

The pediatric cohort construction is anchored on the paper's own reported
values, which gives a free consistency check: the LIBRETTO-121 median weight of
48.5 kg at a median age of 14.0 years must, through Mosteller, reproduce the
reported median BSA of 1.47 m^2.

```{r bsa-anchor}
bsa_at_median <- mosteller(wt_for_age(14) * wt_scale, ht_for_age(14))
c(bsa_mosteller = round(bsa_at_median, 3), bsa_reported_table1 = 1.47)
stopifnot(abs(bsa_at_median - 1.47) < 0.05)
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod, events = events,
  keep = c("regimen", "AGE", "BSA", "WT", "DOSE")
)

sim_df <- as.data.frame(sim) |>
  dplyr::filter(!is.na(Cc))

# The 1 ng/mL assay limit of quantification (Liu 2025 Methods "Sample
# Collection and Analysis"); BQL data were excluded from the paper's analysis.
sim_df <- sim_df |>
  dplyr::mutate(sim = pmax(sim, 1 / 2))

stopifnot(nrow(sim_df) > 0, !anyNA(sim_df$Cc), !anyNA(sim_df$sim))
```

# Replicate published figures

## Concentration-time profile across the C1D8 window

```{r figure-2-style}
# Companion to Figure 2 of Liu 2025 (prediction-corrected VPC, plotted there
# against time after dose in ng/mL). Shown here as the simulated
# concentration-time envelope across the Cycle 1 Day 8 interval that Figure 2's
# middle panels cover.
sim_df |>
  dplyr::filter(time >= 168, time <= 192) |>
  dplyr::mutate(tad_day8 = time - 168) |>
  dplyr::group_by(regimen, tad_day8) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad_day8, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~regimen) +
  labs(x = "Time within Cycle 1 Day 8 (h)",
       y = "Selpercatinib concentration (ng/mL)",
       title = "Simulated Cycle 1 Day 8 profiles",
       caption = "Median and 5th-95th percentiles. Companion to Figure 2 of Liu 2025.")
```

## Figures 3 and 4: exposure by age against the adult reference

The paper's central result is that the Table 3 regimen matches adult exposure
across ages 2-17 years (Figure 4), whereas adapted flat dosing overshoots it in
younger children (Figure 3, left panels). The next chunk computes the Cycle 1
Day 8 exposures that both figures plot.

```{r c1d8-exposure}
adult_ref <- sim_df |>
  dplyr::filter(regimen == "Adults 160 mg BID", time >= 168, time <= 192) |>
  dplyr::group_by(id) |>
  dplyr::summarise(
    cmax = max(Cc),
    auc24 = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    .groups = "drop"
  )

adult_target <- c(cmax = median(adult_ref$cmax), auc24 = median(adult_ref$auc24))

peds_exposure <- sim_df |>
  dplyr::filter(regimen != "Adults 160 mg BID", time >= 168, time <= 192) |>
  dplyr::group_by(regimen, id, AGE, WT, BSA, DOSE) |>
  dplyr::summarise(
    cmax = max(Cc),
    auc24 = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    .groups = "drop"
  )

# Adult reference level, mg/L and mg*h/L, as plotted in Figures 3 and 4.
c(adult_cmax_mgL = round(adult_target[["cmax"]] / 1000, 2),
  adult_auc24_mghL = round(adult_target[["auc24"]] / 1000, 1))
```

```{r figure-3-4}
# Replicates Figures 3 and 4 of Liu 2025: simulated Cycle 1 Day 8 Cmax and
# AUC0-24 by age at baseline, with the adult reference level overlaid. Liu 2025
# plots both on the mg/L and mg*h/L scales, followed here.
peds_exposure |>
  tidyr::pivot_longer(c(cmax, auc24), names_to = "metric", values_to = "value") |>
  dplyr::mutate(
    value = value / 1000,
    metric = factor(metric, c("cmax", "auc24"),
                    c("C1D8 Cmax (mg/L)", "C1D8 AUC0-24 (mg*h/L)")),
    reference = ifelse(metric == "C1D8 Cmax (mg/L)",
                       adult_target[["cmax"]] / 1000,
                       adult_target[["auc24"]] / 1000)
  ) |>
  ggplot(aes(AGE, value)) +
  geom_point(alpha = 0.35, size = 0.8) +
  geom_hline(aes(yintercept = reference), colour = "darkorange",
             linetype = "dashed", linewidth = 0.8) +
  geom_smooth(se = FALSE, colour = "steelblue", method = "loess",
              formula = y ~ x) +
  facet_grid(metric ~ regimen, scales = "free_y") +
  labs(x = "Age at baseline (years)", y = NULL,
       title = "Simulated Cycle 1 Day 8 exposure by age",
       caption = paste("Replicates Figures 3 and 4 of Liu 2025.",
                       "Dashed line: simulated adult reference level at 160 mg BID."))
```

```{r figure-4-gate}
# The paper's conclusion, as an assertion. Both sides are simulated cohorts
# whose spread is driven by which subjects were drawn, so the gates are on the
# CENTRE and on robust quantiles -- never on the extremes (see CLAUDE.md).
rec <- dplyr::filter(peds_exposure, regimen == "Recommended (Table 3)")
flat <- dplyr::filter(peds_exposure, regimen == "Adapted flat dosing")

exposure_ratios <- c(
  rec_auc = median(rec$auc24) / adult_target[["auc24"]],
  rec_cmax = median(rec$cmax) / adult_target[["cmax"]],
  flat_auc_under6 = median(flat$auc24[flat$AGE < 6]) / adult_target[["auc24"]],
  rec_auc_under6 = median(rec$auc24[rec$AGE < 6]) / adult_target[["auc24"]]
)
round(exposure_ratios, 3)

stopifnot(
  # Table 3 regimen matches adult exposure over the whole 2-17 year range.
  abs(exposure_ratios[["rec_auc"]] - 1) < 0.25,
  abs(exposure_ratios[["rec_cmax"]] - 1) < 0.25,
  # Adapted flat dosing overshoots in young children (Figure 3, left panels:
  # "C max and AUC 0-24 values on C1D8 would exceed the exposure observed in
  # adult patients"), and the Table 3 regimen fixes that overshoot.
  exposure_ratios[["flat_auc_under6"]] > 1.5,
  exposure_ratios[["flat_auc_under6"]] > exposure_ratios[["rec_auc_under6"]],
  abs(exposure_ratios[["rec_auc_under6"]] - 1) < 0.35
)
```

The packaged model reproduces the paper's conclusion quantitatively: under the
Table 3 regimen the median pediatric C1D8 AUC0-24 and Cmax land within a couple
of percent of the simulated adult reference, including in the youngest children,
while the adapted flat regimen the paper rejected delivers roughly twice the
adult AUC0-24 to children under 6 years.

# PKNCA validation

NCA is run on `Cc` rather than on `sim`, because the exposures Liu 2025 plots in
Figures 3 and 4 are model-derived simulated values, not observations carrying
residual error; a Cmax taken from `sim` is upward-biased relative to a
model-derived reference. The interval is the Cycle 1 Day 8 window, 168-192 h,
which is where both figures are read.

```{r pknca}
sim_nca <- sim_df |>
  dplyr::select(id, time, Cc, regimen)

# Guarantee a record at the interval start (168 h) and at time 0 so PKNCA has
# an anchor and does not warn about an AUC range starting before the first
# measurement.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, regimen) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, regimen, time, .keep_all = TRUE) |>
  dplyr::arrange(id, regimen, time)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(sim_nca), Cc ~ time | regimen + id)

dose_df <- events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, regimen) |>
  as.data.frame()

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | regimen + id)

intervals <- data.frame(
  start     = 168,
  end       = 192,
  cmax      = TRUE,
  tmax      = TRUE,
  auclast   = TRUE,
  cav       = TRUE,
  half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))

# `half.life = TRUE` also returns the lambda-z diagnostics PKNCA needs to
# compute it (lambda.z, r.squared, span.ratio, ...); keep only the reported
# parameters. Column selection is by NAME, so the order pivot_wider happens to
# emit is irrelevant.
nca_summary <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "cav", "half.life")) |>
  dplyr::group_by(regimen, PPTESTCD) |>
  dplyr::summarise(median = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median)

nca_summary |>
  dplyr::select(regimen, cmax, tmax, auclast, cav, half.life) |>
  dplyr::mutate(
    cmax = round(cmax / 1000, 2), auclast = round(auclast / 1000, 1),
    cav = round(cav / 1000, 2), tmax = round(tmax, 2),
    half.life = round(half.life, 1)
  ) |>
  dplyr::rename(
    "Regimen"              = regimen,
    "Cmax (mg/L)"          = cmax,
    "Tmax within C1D8 (h)" = tmax,
    "AUC0-24 (mg*h/L)"     = auclast,
    "Cavg (mg/L)"          = cav,
    "t1/2 (h)"             = half.life
  ) |>
  knitr::kable(caption = "PKNCA on the Cycle 1 Day 8 interval (168-192 h), per-subject medians.")
```

## Comparison against published exposure

Liu 2025 reports no NCA table. The only published exposure numbers are the adult
reference levels drawn as the orange dashed line in Figures 3 and 4, which are
read off the figures here (roughly 3 mg/L for C1D8 Cmax and 53 mg*h/L for C1D8
AUC0-24 at 160 mg twice daily). The AUC reference is independently corroborated
by the closed form: `2 * 160 / 6.04 = 53.0 mg*h/L` over the two dosing intervals
that make up a day, matching the plotted line to the precision the figure can be
read at.

```{r nca_compare}
published <- tibble::tribble(
  ~regimen,               ~cmax,  ~auclast,
  "Adults 160 mg BID",    3000,   53000
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "regimen",
  units         = c(cmax = "ng/mL", auclast = "ng*h/mL", tmax = "h",
                    cav = "ng/mL", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste("Simulated vs published adult reference exposure.",
                  "* differs from the reference by >20%.",
                  "Reference values are read off the Figure 3 / 4 dashed line."),
  align = c("l", "l", "r", "r", "r")
)
```

```{r nca_compare_gate}
adult_nca <- as.data.frame(nca_res) |>
  dplyr::filter(regimen == "Adults 160 mg BID") |>
  dplyr::group_by(PPTESTCD) |>
  dplyr::summarise(median = median(PPORRES), .groups = "drop")

adult_med <- setNames(adult_nca$median, adult_nca$PPTESTCD)

# Gate on the CENTRE of the adult arm against the figure-read reference. The
# AUC bound is the tighter of the two because the closed form corroborates the
# reference value; Cmax depends on the absorption parameters and on the drawn
# weight distribution, so it gets the wider envelope.
stopifnot(
  abs(adult_med[["auclast"]] / 53000 - 1) < 0.15,
  abs(adult_med[["cmax"]] / 3000 - 1) < 0.25,
  # Terminal half-life is not published, but must be consistent with a regimen
  # the paper dosed twice daily and describes as at steady state by C1D8.
  adult_med[["half.life"]] > 8, adult_med[["half.life"]] < 60
)
```

# Assumptions and deviations

* **Race distribution.** Liu 2025 Table 1 reports sex, age, weight and BSA but
  no race breakdown, even though Asian race on F1 is in the final model. All
  simulated arms are therefore set to `RACE_ASIAN = 0` (the reference category),
  and the covariate's effect is verified separately as exact algebra in the
  covariate-regression chunk above (F1 = 1.183 for `RACE_ASIAN = 1`) rather than
  by sampling a prevalence that the paper does not publish.
* **Pediatric age-weight-height mapping.** The model reads weight, but the
  Table 3 dose bands are defined on BSA, so a height is needed to compute
  Mosteller BSA. Liu 2025 sampled covariates from the NHANES DXA database, which
  is not reproduced here. Instead age is drawn uniformly over 2-18 years (to
  cover the Figure 4 x-axis evenly) and weight and height are read off
  50th-percentile paediatric growth references, with the weight curve rescaled
  by a single factor so that it passes through the paper's own LIBRETTO-121
  median pair (14.0 years, 48.5 kg; Table 1). The resulting median BSA at that
  age reproduces the reported 1.47 m^2 to within 0.05 m^2, which is asserted
  above. Weight carries 15% and height 4% lognormal scatter.
* **Adult weight distribution.** Drawn lognormal to the LIBRETTO-001 median of
  67.3 kg with the Table 1 spread, truncated to the observed 26.8-179 kg range.
  The paper's adult reference line comes from NHANES adults instead, whose
  weight distribution is heavier; this is the main reason the simulated adult
  Cmax sits slightly below the figure-read 3 mg/L.
* **Dose covariate.** `DOSE` is set equal to the per-administration amount on
  every record and held constant, i.e. no dose reductions or escalations are
  simulated. Liu 2025 modelled `DOSE` as time-varying precisely to absorb the
  reductions that occurred in the trials, but it does not publish the reduction
  pattern, so a constant dose per subject is the only reproducible choice.
* **40 mg three times a day is simulated as every 8 hours.** Table 3 says "TID"
  without specifying the clock times.
* **Cycle 1 Day 8 window.** Taken as 168-192 h with the first dose at time 0
  (i.e. Day 1 spans 0-24 h). Figures 3 and 4 report "Cycle 1 Day 8" exposures
  without stating the exact window.
* **AUC0-24 covers two dosing intervals**, as Liu 2025 defines it ("area under
  the curve over two dosing intervals (AUC0-24)"). For the 40 mg three-times-a-
  day arm the same 24 h window contains three doses; the paper plots that arm on
  the same AUC0-24 axis, so the window and not the dose count is what is held
  fixed.
* **No parameter came from outside the paper.** Every `ini()` value is from
  Liu 2025 Table 2. No author correspondence, no upstream model, no figure
  digitisation for any parameter value. The only digitised numbers anywhere in
  this vignette are the two adult reference levels (3 mg/L and 53 mg*h/L) read
  off the Figure 3 / 4 dashed line, which are validation targets and feed no
  parameter.
* **No maturation term.** The model has none, and the paper is explicit that
  predictions below 2 years of age may overpredict concentrations because
  CYP3A4, selpercatinib's main metabolic route, matures around 2 years and no
  patient under 2 years was in the dataset. Simulations here start at 2 years,
  matching the approved age range.

# Errata and internal discrepancies in the source

* **The additive residual error's unit tag is wrong.** Table 2 prints "Additive
  RUV (mg/L) 61.5", but 61.5 mg/L cannot be a concentration standard deviation
  for this model: Figure 3 puts the adult C1D8 Cmax at 160 mg twice daily near
  3 mg/L, so the tag would make the additive error roughly twenty times the peak
  concentration. The value is on the analysis-dataset scale of **ng/mL**, which
  Figure 2 states on its y-axis ("Prediction-corrected selpercatinib
  concentrations (ng/mL)", data spanning about 1000-10000 ng/mL) and which
  matches the 1 ng/mL assay limit of quantification. 61.5 ng/mL is 2.5% of the
  roughly 2500 ng/mL median steady-state concentration. The model file is
  written on the ng/mL scale so that the Table 2 number is used verbatim.
* **Two concentration scales coexist in the paper.** Figure 2 (the pcVPC, i.e.
  the fitting scale) is in ng/mL while Figures 3 and 4 (the simulated exposures)
  are in mg/L and mg*h/L. Both are used above: the model observes in ng/mL and
  the figure replications divide by 1000.
* **The IIV convention is not stated.** Table 2's IIV column gives percentages
  with no footnote saying whether they are `sqrt(exp(omega^2) - 1) * 100` or
  `omega * 100`, and the column carries no confidence interval that could settle
  it by variance symmetry. The log-normal reading `omega^2 = log(CV^2 + 1)` is
  used here. Under the alternative `omega = CV` reading the standard deviations
  would be larger by 6% (CL/F 0.488 vs 0.462), 10% (Vc/F 0.661 vs 0.602), 9%
  (ka 0.636 vs 0.583) and 7% (Dur 0.562 vs 0.524). None of the assertions above
  is sensitive to the choice, because all of them are on medians, on
  typical-value algebra, or on ratios between two arms that share the same
  omegas.
* **Both residual terms are read as standard deviations, not variances.**
  Table 2 tags them with linear scales -- a concentration unit for the additive
  term and "fraction" for the proportional one, where a variance would carry
  squared units -- and the same table reports IIV as CV%, i.e. on the
  standard-deviation scale throughout. Under a variance reading the proportional
  error would be 50.3% rather than 25.3% and the additive term negligible.
* **The Table 2 dose effect on CL/F is linear and does not extrapolate.** The
  multiplier `1 - 0.00321 * (DOSE - 160)` reaches zero at 471.5 mg and turns
  negative above it. No clamp is imposed in the model file because the paper
  imposes none; keep `DOSE` inside the 20-240 mg range the model was fit to.
* **A statistically significant age effect on CL/F was deliberately excluded.**
  Automated stepwise covariate modelling retained age on CL/F after backward
  deletion (p < 0.001), and the Discussion reports that adding it dropped the
  objective function by 43.998 points as a linear model centred on the median
  age of 58 years and by 49.995 points as a segmented "hockey stick" model with
  the break point at the same age. Both, however, "resulted in biased population
  predictions in children" and underpredicted the C1D8 profiles in
  LIBRETTO-121, so the effect was removed during model refinement. Neither
  coefficient is published, so neither is encodable; the excluded covariate is
  documented in the model file's `covariatesDataExcluded$AGE`.
* **Figure 1's caption is garbled in the typeset article.** The caption text
  reads "Schematic view of the final population is the covariate associated
  pharmacokinetic model. Note: Text in italics with the adjacent pharmacokinetic
  parameter", i.e. two sentences interleaved. The intended reading, recoverable
  from the layout, is "Schematic view of the final population pharmacokinetic
  model. Note: Text in italics is the covariate associated with the adjacent
  pharmacokinetic parameter." The schematic itself is unambiguous.
* **Table 1's BSA range is printed without a separator** ("0.4462.79" in the
  Total column), which is 0.446-2.79 as the study columns show.
* **Table S1 double-counts a CYP3A4 inducer row.** The strong and moderate
  CYP3A4 inducer rows each report 3 (0.4%) patients in LIBRETTO-001 and 0 in
  LIBRETTO-121, yet both give a Total of 6 (0.4%). The "Any CYP3A4 inducers"
  row (6 patients) is the consistent one. This affects no model parameter --
  concomitant medication was screened and not retained.
