---
title: "Hydroxychloroquine in cynomolgus macaques (Liu 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hydroxychloroquine in cynomolgus macaques (Liu 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Liu_2020_hydroxychloroquine_monkey"))$reference`
* Description: `r rxode2::rxode(readModelDb("Liu_2020_hydroxychloroquine_monkey"))$description`
* Article: <https://doi.org/10.3389/fphar.2020.602880> (open access; PMC7841297)

**Read this first: two of the paper's printed artefacts do not match the
model's own predictions.** Table 4 and Equations 1-8 are encoded exactly as
printed. With them the model reproduces the observed single-dose (3 mg/kg)
whole-blood and plasma NCA reported in the Results. It also reproduces, to
within 3%, the paper's own population predictions for the lung samples of
groups E and F (Figure 5B). It does not reproduce the simulated Figure 9 and
Table 5. There the simulated lung concentrations sit a uniform ~2-fold above
Table 5, although their terminal decline matches. The Figure 9 regimen is not
stated, but the lung plateau is saturable and almost independent of dose, so
no choice of regimen closes the gap. Separately, the lung concentrations the
model was fitted to (Figure 5) are 8- to 11-fold higher than the lung means
printed in Table 1. The paper's own tissue-to-blood and tissue-to-plasma
ratios agree with the Figure 5 scale, not with Table 1. Details are in the
sections below, and the gates hold these findings so that a later change to
the model shows up.

# Population

Seventeen male cynomolgus macaques (4.13 +/- 0.43 kg) received hydroxychloroquine
intragastrically in six regimens between 1 and 21 mg/kg, over 1 to 5 days
(Methods, "Pharmacokinetics Analysis" and "Subjects and Study Design"). Doses
given on the same day were 4 h apart. The model was fitted jointly to 141
plasma samples, 149 whole-blood samples and one terminal lung sample in 14 of
the 17 animals, taken at necropsy between 120 and 504 h after the first dose.
The same information is available programmatically:

```{r population}
str(readModelDb("Liu_2020_hydroxychloroquine_monkey")()$population)
```

# Source trace

Every `ini()` value carries an in-file comment pointing to its source. The
table below collects them.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Central (plasma) amount, Equation 1 | `d/dt(central)` | Equation 1; Figure 2 |
| Depot, Equation 2 | `d/dt(depot) = -ka * depot` | Equation 2 |
| Red-cell compartment, Equation 3 | `d/dt(rbc_hcq)` | Equation 3; Figure 2 |
| Lung compartment, Equation 4 | `d/dt(lung)` | Equation 4; Figure 2 |
| Peripheral compartment, Equation 5 | `d/dt(peripheral1)` | Equation 5 |
| Plasma / whole-blood / lung concentration | `Cc`, `Cblood`, `Clung` | Equations 6-8 |
| `lka` (Ka) | log(0.592 1/h) | Table 4 |
| `lkel` (Ke) | log(0.236 1/h) | Table 4 |
| `lvc` (Vc/F) | log(114 L) | Table 4 |
| `lk12` (Kcp) | log(0.600 1/h) | Table 4 |
| `lk21` (Kpc) | log(0.514 1/h) | Table 4 |
| `lvmax_rbc` (Kcbmax) | log(2.48 mg/h) | Table 4 (unit: see Assumptions) |
| `lkm_rbc` (Acb50) | log(7.05 mg) | Table 4 |
| `lkeff_rbc` (Kbc) | log(0.718 1/h) | Table 4 |
| `lv_rbc` (Vb/F) | log(2.68 L) | Table 4 |
| `lvmax_lung` (Kclmax) | log(1.92 mg/h) | Table 4 (unit: see Assumptions) |
| `lkm_lung` (Acl50) | log(0.498 mg) | Table 4 |
| `lk_lung_central` (Klc) | log(0.159 1/h) | Table 4 |
| `lv_lung` (VL/F) | log(5.55 L) | Table 4 |
| 13 `eta` variances | log(1 + CV^2) from the IIV CV% column | Table 4 |
| `propSd`, `propSd_Cblood`, `propSd_Clung` | 0.286, 0.118, 0.0514 | Table 4 (PD pl, PD bl, PD lu) |
| `addSd`, `addSd_Cblood`, `addSd_Clung` | 1e-5, fixed | Table 4 (SD pl, SD bl, SD lu) |
| Residual form | proportional plus additive (`combined1`) | Methods, "Structural PK Model" and "Software and Platform Used" |

# Model setup

```{r model}
mod <- rxode2::rxode2(readModelDb("Liu_2020_hydroxychloroquine_monkey"))
mod_typ <- rxode2::zeroRe(mod)
# The explicit ODEs must be solved as written (no linear-compartment
# auto-solve replacing them).
stopifnot(is.null(mod$linCmt))

wt <- 4.13 # kg, mean body weight (Methods)
```

Doses are given in mg as mg/kg multiplied by the 4.13 kg mean body weight. All
observation records sit on the `central` state with `dvid = 1`. The model
declares three error endpoints, so every observation row has to name one, and
`Cc`, `Cblood` and `Clung` are all returned as columns at each of those
records.

```{r helpers}
# One event table per regimen. `doses` is a data frame of time (h) and
# dose (mg/kg); observations are on the central state.
make_events <- function(doses, obs_times, id = 1L, regimen = "") {
  dose_rows <- data.frame(
    id = id, time = doses$time, amt = doses$mgkg * wt,
    evid = 1L, cmt = "depot", dvid = NA_integer_
  )
  obs_rows <- data.frame(
    id = id, time = obs_times, amt = 0, evid = 0L, cmt = "central", dvid = 1L
  )
  out <- dplyr::bind_rows(dose_rows, obs_rows) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
  out$regimen <- regimen
  out
}

# rxSolve() omits the `id` column when only one subject is solved; restore it.
solve_df <- function(model, events) {
  out <- as.data.frame(rxode2::rxSolve(model, events, keep = "regimen"))
  if (!"id" %in% names(out)) out$id <- 1L
  out
}

regimens <- list(
  "3 mg/kg single" = data.frame(time = 0, mgkg = 3),
  "3 mg/kg x2 on days 1-2" = data.frame(time = c(0, 4, 24, 28), mgkg = 3),
  "2 mg/kg x2 then 1 mg/kg x2 daily" = data.frame(
    time = c(0, 4, 24, 28, 48, 52), mgkg = c(2, 2, 1, 1, 1, 1)
  ),
  "6 mg/kg x2 then 2 mg/kg x2 daily" = data.frame(
    time = c(0, 4, 24, 28, 48, 52), mgkg = c(6, 6, 2, 2, 2, 2)
  ),
  "21 mg/kg x2 (group E)" = data.frame(time = c(0, 4), mgkg = 21),
  "21 mg/kg x2 then 7 mg/kg x2 daily (group F)" = data.frame(
    time = c(0, 4, 24, 28, 48, 52, 72, 76, 96, 100),
    mgkg = c(21, 21, rep(7, 8))
  )
)
```

# Replicate Figure 1: observed-design regimens

Figure 1 of the paper shows the observed whole-blood and plasma concentrations
for the six regimens. The typical-value predictions for the same designs are
below. The sampling windows are 72 h for the four low-dose groups, 264 h for
group E and 120 h for group F.

```{r fig1, fig.width = 8, fig.height = 5}
end_time <- c(72, 72, 72, 72, 264, 120)
ev_fig1 <- dplyr::bind_rows(lapply(seq_along(regimens), function(i) {
  make_events(
    regimens[[i]], seq(0, end_time[i], by = 0.5),
    id = i, regimen = names(regimens)[i]
  )
}))
sim_fig1 <- solve_df(mod_typ, ev_fig1)

# Drop the pre-dose zero so it does not break the log axis.
sim_fig1[sim_fig1$time != 0, ] |>
  tidyr::pivot_longer(c(Cc, Cblood), names_to = "matrix", values_to = "conc") |>
  dplyr::mutate(matrix = dplyr::recode(matrix, Cc = "Plasma", Cblood = "Whole blood")) |>
  ggplot(aes(time, conc, colour = regimen)) +
  geom_line() +
  facet_wrap(~matrix) +
  scale_y_log10() +
  labs(
    x = "Time after first dose (h)", y = "HCQ concentration (ng/mL)",
    colour = NULL,
    caption = "Typical-value predictions for the regimens of Figure 1 of Liu 2020."
  ) +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(ncol = 2))
```

# Single-dose NCA against the observed data

The Results report the NCA from WinNonlin after a single 3 mg/kg dose, with
blood and plasma sampled at 0, 1, 2, 4, 8, 24, 48 and 72 h. Cmax was
292.33 +/- 114.66 ng/mL in blood and 36.90 +/- 22.52 ng/mL in plasma. AUC0-inf
was 5,978.94 +/- 1,981.30 h\*ng/mL in blood and 363.31 +/- 195.38 h\*ng/mL in
plasma. The simulated profiles are put through PKNCA on the same sampling
times, so AUC0-inf is extrapolated from a 72 h profile in the same way as the
observed value.

```{r nca-sim}
nca_times <- c(0, 1, 2, 4, 8, 24, 48, 72)
sd_dose <- regimens[["3 mg/kg single"]]

# Typical-value animal (deterministic).
ev_typ <- make_events(sd_dose, nca_times, id = 1L, regimen = "3 mg/kg single")
sim_typ <- solve_df(mod_typ, ev_typ)

# Stochastic cohort for the between-animal spread (100 animals).
rxode2::rxSetSeed(20201116)
n_anim <- 100L
ev_coh <- dplyr::bind_rows(lapply(seq_len(n_anim), function(i) {
  make_events(sd_dose, nca_times, id = i, regimen = "3 mg/kg single")
}))
sim_coh <- solve_df(mod, ev_coh)
```

```{r nca-run}
run_nca <- function(sim, conc_col) {
  conc_df <- sim |>
    dplyr::transmute(id, time, treatment = regimen, conc = .data[[conc_col]]) |>
    dplyr::filter(!is.na(conc))
  dose_df <- data.frame(
    id = unique(conc_df$id), time = 0, treatment = "3 mg/kg single",
    amt = 3 * wt
  )
  conc_obj <- PKNCA::PKNCAconc(conc_df, conc ~ time | treatment + id)
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)
  intervals <- data.frame(
    start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
    auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE
  )
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
  as.data.frame(res$result)
}

nca_typ <- dplyr::bind_rows(
  run_nca(sim_typ, "Cc") |> dplyr::mutate(matrix = "Plasma"),
  run_nca(sim_typ, "Cblood") |> dplyr::mutate(matrix = "Whole blood")
)
nca_coh <- dplyr::bind_rows(
  run_nca(sim_coh, "Cc") |> dplyr::mutate(matrix = "Plasma"),
  run_nca(sim_coh, "Cblood") |> dplyr::mutate(matrix = "Whole blood")
)
```

```{r nca-compare}
observed <- data.frame(
  matrix = c("Plasma", "Whole blood"),
  cmax = c(36.90, 292.33),
  aucinf.obs = c(363.31, 5978.94)
)

cmp_typ <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_typ, reference = observed, by = "matrix",
  params = c("cmax", "aucinf.obs"),
  units = c(cmax = "ng/mL", aucinf.obs = "h*ng/mL")
)
cmp_coh <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_coh, reference = observed, by = "matrix",
  params = c("cmax", "aucinf.obs"),
  units = c(cmax = "ng/mL", aucinf.obs = "h*ng/mL")
)

knitr::kable(
  cmp_typ,
  caption = "Typical-value animal vs. observed mean (n = 3), 3 mg/kg single dose."
)
knitr::kable(
  cmp_coh,
  caption = "Median of 100 simulated animals vs. observed mean (n = 3), 3 mg/kg single dose."
)
```

`r paste(unique(c(attr(cmp_typ, "footnote"), attr(cmp_coh, "footnote"))), collapse = " ")`

The median of the simulated cohort reproduces both observed AUC0-inf values
and the whole-blood Cmax. Its plasma Cmax sits below the observed mean, but
well inside the observed +/- 1 SD range (14.4 to 59.4 ng/mL). An observed mean
of three animals, with a 61% CV on Cmax, is a weak reference for any single
number.

The typical-value animal is not the cohort median here. The red-cell and lung
transfers saturate, and their parameters carry CVs above 100%, so the median
of the simulated profiles sits away from the profile of the median animal.
The typical animal's terminal phase is also slower than the observed one:
extrapolation beyond 72 h makes up about a quarter of its whole-blood AUC0-inf,
where the paper reports under 10% for blood and under 20% for plasma. Its
AUC0-inf therefore overshoots. Its AUC0-72 needs no extrapolation. For plasma
it lies inside the range implied by the reported AUC0-inf and extrapolated
share, and for whole blood it lies about 5% above that range.

```{r nca-gate}
typ_val <- function(mat, code) {
  v <- nca_typ$PPORRES[nca_typ$matrix == mat & nca_typ$PPTESTCD == code]
  if (length(v) != 1L) stop("no unique NCA value for ", mat, " ", code)
  v
}
coh_med <- function(mat, code) {
  v <- nca_coh$PPORRES[nca_coh$matrix == mat & nca_coh$PPTESTCD == code]
  if (length(v) != n_anim) stop("expected one value per animal for ", mat, " ", code)
  median(v, na.rm = TRUE)
}
# Observed AUC0-72 implied by the reported AUC0-inf and the reported
# extrapolated share (< 20% plasma, < 10% blood; Results).
auc72_range <- rbind(
  Plasma = c(0.80, 1.00) * 363.31,
  `Whole blood` = c(0.90, 1.00) * 5978.94
)
gate_tab <- data.frame(
  Quantity = c(
    "Typical: plasma Cmax (ng/mL)", "Typical: whole-blood Cmax (ng/mL)",
    "Typical: plasma AUC0-72 (h*ng/mL)", "Typical: whole-blood AUC0-72 (h*ng/mL)",
    "Typical: whole-blood extrapolated share of AUC0-inf (%)",
    "Cohort median: plasma AUC0-inf (h*ng/mL)", "Cohort median: whole-blood AUC0-inf (h*ng/mL)"
  ),
  Simulated = c(
    typ_val("Plasma", "cmax"), typ_val("Whole blood", "cmax"),
    typ_val("Plasma", "auclast"), typ_val("Whole blood", "auclast"),
    100 * (1 - typ_val("Whole blood", "auclast") / typ_val("Whole blood", "aucinf.obs")),
    coh_med("Plasma", "aucinf.obs"), coh_med("Whole blood", "aucinf.obs")
  ),
  Observed = c(
    "36.90 +/- 22.52", "292.33 +/- 114.66",
    "290.6 to 363.3 (implied)", "5381 to 5979 (implied)", "< 10",
    "363.31 +/- 195.38", "5978.94 +/- 1981.30"
  )
)
knitr::kable(gate_tab, digits = 1, caption = "Gated quantities, 3 mg/kg single dose.")

stopifnot(
  # Deterministic typical-value animal. A mis-transcribed volume, rate or
  # unit moves these by tens of percent.
  typ_val("Plasma", "cmax") > 36.90 - 22.52,
  typ_val("Plasma", "cmax") < 36.90 + 22.52,
  abs(typ_val("Whole blood", "cmax") / 292.33 - 1) < 0.15,
  typ_val("Plasma", "auclast") > 0.9 * auc72_range["Plasma", 1],
  typ_val("Plasma", "auclast") < 1.1 * auc72_range["Plasma", 2],
  typ_val("Whole blood", "auclast") > 0.9 * auc72_range["Whole blood", 1],
  typ_val("Whole blood", "auclast") < 1.1 * auc72_range["Whole blood", 2],
  # Cohort median (100 animals): the centre, with headroom for the draw.
  abs(coh_med("Plasma", "aucinf.obs") / 363.31 - 1) < 0.25,
  abs(coh_med("Whole blood", "aucinf.obs") / 5978.94 - 1) < 0.25
)
```

# Lung: population predictions (Figure 5B) and Table 1

The lung data are single terminal samples from 14 animals. Figure 5B plots each
observed lung concentration against the model's population prediction (PRED).
Groups E and F are the only groups necropsied at 264 h and 120 h, so their
points can be identified unambiguously. The maintainers digitised the
circle centres of Figure 5B from the published raster.

```{r lung-fig5b}
fig5b <- data.frame(
  group = c("F", "F", "E", "E", "E"),
  time = c(120, 120, 264, 264, 264),
  pred = c(1968.4, 2022.5, 167.3, 178.2, 181.5),
  obs = c(4875.4, 7731.1, 179.4, 348.7, 152.7)
)
lung_at <- function(reg, t_obs) {
  s <- sim_fig1[sim_fig1$regimen == reg & sim_fig1$time == t_obs, ]
  if (nrow(s) != 1L) stop("no unique row for ", reg, " at ", t_obs, " h")
  s
}
sF <- lung_at("21 mg/kg x2 then 7 mg/kg x2 daily (group F)", 120)
sE <- lung_at("21 mg/kg x2 (group E)", 264)

lung_tab <- fig5b |>
  dplyr::group_by(group, time) |>
  dplyr::summarise(
    fig5b_pred = mean(pred), fig5b_obs = mean(obs), .groups = "drop"
  ) |>
  dplyr::mutate(
    sim_typical = c(E = sE$Clung, F = sF$Clung)[group],
    table1_mean = c(E = 20.77, F = 738.50)[group],
    sim_over_pred = sim_typical / fig5b_pred,
    fig5b_obs_over_table1 = fig5b_obs / table1_mean
  )
lung_tab |>
  dplyr::rename(
    Group = group, `Time (h)` = time,
    `Figure 5B PRED (ng/g)` = fig5b_pred,
    `Figure 5B observed (ng/g)` = fig5b_obs,
    `Simulated typical (ng/g)` = sim_typical,
    `Table 1 mean (ng/g)` = table1_mean,
    `Simulated / PRED` = sim_over_pred,
    `Figure 5B observed / Table 1` = fig5b_obs_over_table1
  ) |>
  knitr::kable(digits = 2, caption = "Lung: typical-value prediction vs. Figure 5B (digitised) and Table 1.")
```

The typical-value animal reproduces the paper's population predictions for
both groups. That confirms the transcription of the lung parameters, the
saturable-transfer form and the concentration scale. The paper's PRED
underpredicts the group F observations about 3-fold, and so does this model.
That is a property of the published fit, not of this implementation.

The observed values in Figure 5 are 8- to 11-fold higher than the means in
Table 1. The Methods homogenise each tissue with ten volumes of 50% methanol,
and the Table 1 values look like homogenate concentrations that were not
corrected for that dilution. Two things support this. The Kp ratios printed
next to them in Tables 1 and 3 are consistent with the Figure 5 scale. And
10 x 738.50 ng/g divided by the ~27 ng/mL group F plasma level read from
Figure 1C gives about 270, against the printed lung-to-plasma Kp of 267.38.
This is an interpretation; the paper does not state it. The model is compared
here with Figure 5, the data it was fitted to.

```{r lung-gate}
# Deterministic typical-value predictions against the digitised PRED.
# Realised: 1.00 (group F) and 1.03 (group E). Digitising error is about 2%.
stopifnot(
  nrow(lung_tab) == 2L,
  all(abs(lung_tab$sim_over_pred - 1) < 0.08),
  # The documented Table 1 scale gap.
  all(lung_tab$fig5b_obs_over_table1 > 7), all(lung_tab$fig5b_obs_over_table1 < 12)
)
```

# Replicate Figure 9 and Table 5: simulated multiple-dose profile

The paper simulates "one dosing regimen" with the typical parameters, and
Table 5 reports concentrations at 12 to 96 h after the last dose. The regimen
is not stated. Figure 9 shows two 12-hourly loading doses and then eight
12-hourly maintenance doses (last dose at 108 h). The regimen below uses group
F's dose levels (21 mg/kg, then 7 mg/kg) on that 12-hourly schedule. This is an
assumption for illustration only.

```{r fig9, fig.width = 8, fig.height = 3.5}
fig9_doses <- data.frame(time = seq(0, 108, by = 12), mgkg = c(21, 21, rep(7, 8)))
tab5 <- data.frame(
  tad = c(12, 24, 48, 72, 96),
  Cc = c(10.63, 4.26, 1.80, 1.10, 0.76),
  Cblood = c(187.29, 85.40, 37.68, 23.19, 16.11),
  Clung = c(1008.54, 865.68, 629.94, 482.23, 380.55)
)
ev9 <- make_events(
  fig9_doses, sort(unique(c(seq(0, 348, by = 0.5), 108 + tab5$tad))),
  regimen = "Figure 9 (assumed)"
)
sim9 <- solve_df(mod_typ, ev9)

sim9 |>
  tidyr::pivot_longer(c(Cc, Cblood, Clung), names_to = "matrix", values_to = "conc") |>
  dplyr::mutate(matrix = factor(
    matrix, c("Cc", "Cblood", "Clung"),
    c("Plasma (ng/mL)", "Whole blood (ng/mL)", "Lung (ng/g)")
  )) |>
  ggplot(aes(time, conc)) +
  geom_line() +
  geom_point(
    data = tab5 |>
      tidyr::pivot_longer(-tad, names_to = "matrix", values_to = "conc") |>
      dplyr::mutate(
        time = 108 + tad,
        matrix = factor(
          matrix, c("Cc", "Cblood", "Clung"),
          c("Plasma (ng/mL)", "Whole blood (ng/mL)", "Lung (ng/g)")
        )
      ),
    colour = "red"
  ) +
  facet_wrap(~matrix, scales = "free_y") +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "HCQ concentration",
    caption = "Line: typical-value simulation, assumed regimen. Red points: Table 5 of Liu 2020."
  )
```

```{r tab5}
sim_tab5 <- sim9[match(108 + tab5$tad, sim9$time), c("Cc", "Cblood", "Clung")]
stopifnot(nrow(sim_tab5) == 5L, !anyNA(sim_tab5))
tab5_cmp <- data.frame(
  `TAD (h)` = tab5$tad,
  `Plasma, Table 5` = tab5$Cc, `Plasma, sim` = sim_tab5$Cc,
  `Blood, Table 5` = tab5$Cblood, `Blood, sim` = sim_tab5$Cblood,
  `Lung, Table 5` = tab5$Clung, `Lung, sim` = sim_tab5$Clung,
  `Lung/plasma, Table 5` = tab5$Clung / tab5$Cc,
  `Lung/plasma, sim` = sim_tab5$Clung / sim_tab5$Cc,
  check.names = FALSE
)
knitr::kable(tab5_cmp, digits = 2, caption = "Table 5 of Liu 2020 vs. the typical-value simulation (assumed regimen).")
```

The lung level sits about 2-fold above Table 5 at every time. The lung
decline over the same window matches: from 12 to 96 h after the last dose, the
simulated lung concentration falls by the same factor as in Table 5.
The saturable lung uptake makes the lung plateau `Kclmax / (Klc * VL)`, which is
2,176 ng/g with the printed values, and the simulation approaches it during
maintenance dosing. Figure 9 shows a plateau of about 1,050 ng/g. The plasma
and whole-blood values depend on the unstated regimen and are not compared
further.

```{r tab5-gate}
lung_ratio <- sim_tab5$Clung / tab5$Clung
decline_sim <- sim_tab5$Clung[1] / sim_tab5$Clung[5]
decline_tab <- tab5$Clung[1] / tab5$Clung[5]
c(lung_ratio = round(lung_ratio, 3), decline_ratio = round(decline_sim / decline_tab, 3))
stopifnot(
  # Terminal lung decline: shape reproduced within 5%.
  abs(decline_sim / decline_tab - 1) < 0.05,
  # Lung level: the documented ~2-fold offset from Table 5.
  all(lung_ratio > 1.9), all(lung_ratio < 2.4),
  # Saturable lung plateau Kclmax / (Klc * VL), in ng/g.
  abs(1000 * 1.92 / (0.159 * 5.55) - 2175.8) < 1
)
```

# Assumptions and deviations

* **Unit of Kcbmax and Kclmax.** Table 4 prints no unit for these two
  parameters, and the Results text gives them as h^-1. Equation 1 and the
  Figure 2 arrows (`Kcbmax/(Acb50 + A1)` multiplying A1) make the transfer flux
  `Kmax * A1 / (A50 + A1)` with A1 in mg, which needs Kmax in mg/h. The
  equation is followed. The only reading consistent with an h^-1 unit is
  `Kmax * A50 * A1 / (A50 + A1)`. For the lung term, that reading halves the
  flux, since Acl50 = 0.498 mg, and moves the lung plateau close to Figure 9.
  For the blood term it predicts a single-dose whole-blood Cmax about 5-fold
  above the observed value. It was therefore rejected. Nothing in the paper
  supports reading the two terms differently.
* **Typical values.** The Table 4 "Mean" column is encoded as the typical
  (median) value. The paper refers to these estimates as the typical-monkey
  values (abstract) and uses the "typical parameters" for its simulations
  (Methods).
* **Between-animal variability.** The IIV CV% is converted to a log-normal
  variance as `omega^2 = log(1 + CV^2)`. ADAPT's MLEM fits a multivariate
  log-normal distribution, but only the CVs are printed, so the covariance
  matrix is diagonal. The Results text says the CVs range "from 29 to 88.7%,
  except for IIV of Kcl (160%)". Table 4 lists five CVs above 100%
  (Acb50, Vb/F, Acl50, Kcbmax, Kcp) and no parameter at 160%. Table 4 is used.
* **Residual error.** ADAPT's proportional-plus-additive model is
  `SD = SDinter + SDslope * Y`, encoded as `combined1`. The additive SDs are
  fixed at 1e-5, as printed, so the error is proportional in practice.
* **Lung concentration units.** The lung amount in mg divided by VL in L
  gives mg/L, which is reported as ng/g assuming a tissue density of 1 g/mL.
* **Dose units.** Doses are in mg as reported. The paper does not say whether
  they are expressed as hydroxychloroquine sulfate (the administered material)
  or as base. Simulations use mg/kg x 4.13 kg, the mean body weight. The
  model has no body-size covariate.
* **Group labels.** The Figure 1 caption and the Methods text label the six
  groups differently. The dose levels and regimen shapes agree, and this
  vignette identifies regimens by dose, not by letter. The Tissue Distribution
  section's group E and F labels are used for the lung data.
* **Figure 9 / Table 5 regimen.** Not stated in the paper. The regimen used
  here is an assumption, and the comparison with Table 5 is limited to the
  lung, whose plateau is set by the saturable uptake and not by the dose.
* **Table 1 lung scale.** The Table 1 tissue concentrations are compared
  only as a documented scale gap. The model is validated against the Figure 5B
  population predictions, which it reproduces.
* **Known deviations.** The printed parameters put the lung concentrations
  about 2-fold above the simulated Table 5 and Figure 9 values. The regimen
  behind those two artefacts is unknown. The paper's own population
  prediction also underpredicts the group F lung observations about 3-fold.
  These deviations are recorded, not tuned away, and are held by the gates
  above.
