---
title: "Deutenzalutamide (Liu 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deutenzalutamide (Liu 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Liu_2024_deutenzalutamide"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1002/cpdd.1477>
* Supplement (Tables S1-S4, Figures S1-S3): <https://www.ebi.ac.uk/europepmc/webservices/rest/PMC11609056/supplementaryFiles>

Deutenzalutamide (development code HC-1119) is a deuterated analogue of
enzalutamide, a second-generation androgen receptor antagonist. Liu 2024
developed the first population pharmacokinetic model for the compound from a
single open-label 3 + 3 dose-escalation Phase Ia trial (NCT03774056) in patients
with metastatic castration-resistant prostate cancer (mCRPC).

The final model is a two-compartment model with first-order absorption and
first-order elimination from the central compartment, with body weight on the
apparent peripheral volume of distribution as the only retained covariate.

# Population

```{r population, echo = FALSE}
pop <- ui$population
tibble::tibble(
  Field = names(pop),
  Value = vapply(pop, function(x) paste(as.character(x), collapse = "; "), character(1))
) |>
  knitr::kable(caption = "Population metadata (Liu 2024 Methods and Table S3).")
```

Twenty-four men with mCRPC contributed 543 deutenzalutamide plasma
concentrations. Subjects were allocated to 40 mg (n = 3), 80 mg (n = 9),
160 mg (n = 9) and 200 mg (n = 3) once-daily oral soft capsules for 12 weeks.
Median (range) baseline age was 69.5 (55-80) years, body weight 64 (46-81) kg
and height 165 (149-178) cm (Liu 2024 Table S3). All renal and hepatic
laboratory values were within the normal range, which is why Liu 2024's
Discussion cautions against extrapolating the model to organ impairment.

The same information is available programmatically via
`readModelDb("Liu_2024_deutenzalutamide")()$population`.

# Source trace

Each `ini()` entry in
`inst/modeldb/specificDrugs/Liu_2024_deutenzalutamide.R` carries an in-file
comment naming its origin. They are collected here for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` (ka) | 1.32 1/h (RSE 16.3%) | Liu 2024 Table 1, "Final model" estimates |
| `lcl` (CL/F) | 0.166 L/h (RSE 3.9%) | Liu 2024 Table 1 |
| `lvc` (V1/F) | 17.5 L, fixed | Liu 2024 Table 1, marked "(fixed)" |
| `lvp` (V2/F at 64 kg) | 58.1 L (RSE 8.3%) | Liu 2024 Table 1 |
| `lq` (CLd/F) | 12.5 L/h, fixed | Liu 2024 Table 1, marked "(fixed)" |
| `e_wt_vp` | 1.83 (RSE 25.6%) | Liu 2024 Table 1, row "BW on V2/F" |
| Body-weight reference | 64.0 kg | Liu 2024 Results (Simulation), "the typical value of body weight (64.0 kg)"; equals the Table S3 median |
| Covariate functional form | `(WT / 64)^e_wt_vp` | Liu 2024 Methods (Covariate Model) Equation 3 |
| IIV form | `P_i = P_hat * exp(eta_i)` | Liu 2024 Methods (Statistical Model) Equation 1 |
| `etalcl` | 0.0274 (RSE 41.6%, shrinkage 2.40%) | Liu 2024 Table 1, omega CL/F |
| `etalvp` | 0.0446 (RSE 38.1%, shrinkage 6.40%) | Liu 2024 Table 1, omega V2/F |
| `etalka` | 0.440 (RSE 40.2%, shrinkage 4.80%) | Liu 2024 Table 1, omega ka |
| `propSd` | sqrt(0.0208) = 0.144 | Liu 2024 Table 1, sigma prop; proportional-only per Results |
| Residual-error form | proportional | Liu 2024 Results, "the residual variability was fitted by proportional model" (the eps_1 term of Equation 2) |
| Two-compartment, first-order absorption and elimination | n/a | Liu 2024 Results (Population Pharmacokinetic Modeling) and Conclusion |

## Reported secondary parameter

Liu 2024 Table 1 reports `t1/2 = 72.9` hours, labelled "elimination
half-life". That is *not* the terminal (beta) half-life of the two-compartment
system; it is the half-life of the central-compartment elimination rate
constant, `log(2) * V1/F / (CL/F)`. Recovering it from the model's own
parameters confirms that `CL/F` and `V1/F` were transcribed correctly and
settles which half-life the table reports.

```{r half-life-identity}
theta <- setNames(ui$theta, names(ui$theta))
cl_typ <- exp(theta[["lcl"]])
vc_typ <- exp(theta[["lvc"]])
vp_typ <- exp(theta[["lvp"]])
q_typ  <- exp(theta[["lq"]])

t_half_k10 <- log(2) * vc_typ / cl_typ

# Terminal (beta) half-life of the two-compartment system, for contrast.
k10 <- cl_typ / vc_typ
k12 <- q_typ / vc_typ
k21 <- q_typ / vp_typ
s   <- k10 + k12 + k21
beta <- 0.5 * (s - sqrt(s^2 - 4 * k21 * k10))
t_half_terminal <- log(2) / beta

tibble::tibble(
  Quantity = c("log(2) * V1/F / (CL/F)", "Terminal (beta) half-life"),
  `Hours`  = c(t_half_k10, t_half_terminal),
  `Liu 2024 Table 1 t1/2` = c(72.9, NA_real_)
) |>
  knitr::kable(digits = 1,
               caption = "The reported 72.9 h is the k10 half-life, not the terminal half-life.")

# Deterministic identity computed from the model's own thetas -- a tight bound
# is correct here (no simulated cohort is involved).
stopifnot(abs(t_half_k10 / 72.9 - 1) < 0.01)
```

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the Table S3 body-weight distribution (median 64 kg, range 46-81 kg) and the
four dose-escalation levels, with 100 virtual subjects per dose arm.

```{r cohort}
# set.seed() seeds R's RNG only. rxode2's simulation RNG is partitioned per
# solver thread, so this cohort is reproducible here and different on a machine
# with a different thread count. Every assertion below is written to hold for
# any cohort the model can produce (see the known-failure-patterns reference).
set.seed(20241202)
rxode2::rxSetSeed(20241202)

n_per_arm <- 100L
doses_mg  <- c(40, 80, 160, 200)
tau       <- 24            # h, once daily (Liu 2024 Methods)
n_doses   <- 84L           # 12 weeks of once-daily dosing
t_last_dose <- tau * (n_doses - 1L)

# Body weight: truncated normal matched to the Table S3 median and range.
# Table S3 reports only median and range, so the SD is an assumption
# (see "Assumptions and deviations").
rtnorm_wt <- function(n, mean = 64, sd = 9, lo = 46, hi = 81) {
  out <- numeric(0)
  while (length(out) < n) {
    draw <- stats::rnorm(2 * n, mean, sd)
    out <- c(out, draw[draw >= lo & draw <= hi])
  }
  out[seq_len(n)]
}

make_cohort <- function(n, dose_mg, id_offset = 0L) {
  subj <- tibble::tibble(
    id        = id_offset + seq_len(n),
    WT        = rtnorm_wt(n),
    treatment = paste0(dose_mg, " mg")
  )

  dosing <- subj |>
    tidyr::crossing(time = tau * seq(0, n_doses - 1L)) |>
    dplyr::mutate(amt = dose_mg, evid = 1L, cmt = "depot")

  # Day 1 profile, weekly troughs, and the Day-84 steady-state profile --
  # the sampling schedule of Liu 2024 Methods (Study Design and Population).
  obs_times <- sort(unique(c(
    seq(0, tau, by = 0.25),
    tau * seq(2, n_doses - 1L, by = 7),
    seq(t_last_dose, t_last_dose + tau, by = 0.25)
  )))

  obs <- subj |>
    tidyr::crossing(time = obs_times) |>
    # cmt is the ODE STATE name, never the observable "Cc".
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")

  dplyr::bind_rows(dosing, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  lapply(seq_along(doses_mg), function(i) {
    make_cohort(n_per_arm, doses_mg[i], id_offset = (i - 1L) * n_per_arm)
  })
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
stopifnot(nrow(dplyr::distinct(events, id)) == n_per_arm * length(doses_mg))
```

# Simulation

```{r simulate}
mod <- readModelDb("Liu_2024_deutenzalutamide")

sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep   = c("treatment", "WT")
) |>
  as.data.frame()

sim_obs <- sim |> dplyr::filter(!is.na(Cc))
stopifnot(nrow(sim_obs) > 0, all(sim_obs$Cc >= 0))

# Typical-value (no between-subject variability) version of the model, used for
# every deterministic check and for the typical-value figures below.
mod_typ <- mod |> rxode2::zeroRe()
```

# Replicate published figures

## Figure 1 - steady-state concentration-time profiles by dose group

Liu 2024 Figure 1 plots the observed concentration-time profiles after multiple
administration of 40-200 mg on linear and semi-logarithmic axes. The panels
below are the model's prediction interval over the final (Day 84) dosing
interval.

```{r figure-1}
ss_window <- sim_obs |>
  dplyr::filter(time >= t_last_dose, time <= t_last_dose + tau) |>
  dplyr::mutate(tad = time - t_last_dose,
                treatment = factor(treatment, levels = paste0(doses_mg, " mg")))

ss_bands <- ss_window |>
  dplyr::group_by(treatment, tad) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot(ss_bands, aes(tad, Q50, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time after dose on Day 84 (h)", y = "Deutenzalutamide (ug/mL)",
       colour = "Dose", fill = "Dose",
       title = "Figure 1A - steady-state profiles (linear)",
       caption = "Replicates Figure 1A of Liu 2024.")

ggplot(ss_bands, aes(tad, Q50, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(x = "Time after dose on Day 84 (h)", y = "Deutenzalutamide (ug/mL)",
       colour = "Dose", fill = "Dose",
       title = "Figure 1B - steady-state profiles (semi-log)",
       caption = "Replicates Figure 1B of Liu 2024.")
```

Dose proportionality is a structural consequence of the linear model; Liu 2024's
exploratory analysis reports that exposure "increased with the increase of
dose". Median steady-state AUC scaled by dose should therefore be constant
across arms.

```{r dose-proportionality}
auc_tau <- function(df) {
  df <- df[order(df$tad), ]
  sum(diff(df$tad) * (utils::head(df$Cc, -1) + utils::tail(df$Cc, -1)) / 2)
}

prop_tbl <- ss_window |>
  dplyr::group_by(treatment, id) |>
  dplyr::group_modify(~ tibble::tibble(auc = auc_tau(.x))) |>
  dplyr::ungroup() |>
  dplyr::mutate(dose = as.numeric(sub(" mg", "", treatment))) |>
  dplyr::group_by(treatment, dose) |>
  dplyr::summarise(`Median AUC0-tau (ug*h/mL)` = median(auc), .groups = "drop") |>
  dplyr::mutate(`AUC / dose (h/L)` = `Median AUC0-tau (ug*h/mL)` / dose)

knitr::kable(prop_tbl, digits = c(0, 0, 1, 3),
             caption = paste("Dose-normalised steady-state exposure across dose",
                             "arms. Each arm draws its own random body weights",
                             "and etas, so the medians carry sampling noise."))
```

The cohort medians above each rest on a different random draw of body weights
and random effects, so they are not the right place to gate dose
proportionality. The deterministic typical-value version is:

```{r dose-proportionality-gate}
prop_events <- dplyr::bind_rows(lapply(seq_along(doses_mg), function(i) {
  subj <- tibble::tibble(id = i, WT = 64, treatment = paste0(doses_mg[i], " mg"))
  dosing <- subj |>
    tidyr::crossing(time = tau * seq(0, n_doses - 1L)) |>
    dplyr::mutate(amt = doses_mg[i], evid = 1L, cmt = "depot")
  obs <- subj |>
    tidyr::crossing(time = seq(t_last_dose, t_last_dose + tau, by = 0.05)) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
  dplyr::bind_rows(dosing, obs) |> dplyr::arrange(id, time, dplyr::desc(evid))
}))

prop_typ <- rxode2::rxSolve(mod_typ, events = prop_events,
                            keep = c("treatment", "WT")) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(tad = time - t_last_dose) |>
  dplyr::group_by(treatment) |>
  dplyr::group_modify(~ tibble::tibble(auc = auc_tau(.x))) |>
  dplyr::ungroup() |>
  dplyr::mutate(dose = as.numeric(sub(" mg", "", treatment)),
                dn = auc / dose)

# Deterministic: the model is linear, so dose-normalised AUC must be identical
# across arms to solver tolerance. Only a mis-scaled dose column can break this.
stopifnot((max(prop_typ$dn) / min(prop_typ$dn) - 1) < 1e-6)
```

## Figure 4 - effect of body weight on the approach to steady state

Liu 2024 Figure 4 simulates 80 mg once daily at different body weights. Its
message is that heavier subjects take *longer* to reach steady state, because
body weight increases `V2/F` and therefore the terminal half-life, while the
eventual steady-state exposure is essentially unchanged. These are typical-value
(no between-subject variability) predictions.

```{r figure-4}
wt_arms <- c(50, 60, 70, 80)

wt_events <- dplyr::bind_rows(lapply(seq_along(wt_arms), function(i) {
  w <- wt_arms[i]
  subj <- tibble::tibble(id = i, WT = w, treatment = paste0(w, " kg"))
  dosing <- subj |>
    tidyr::crossing(time = tau * seq(0, n_doses - 1L)) |>
    dplyr::mutate(amt = 80, evid = 1L, cmt = "depot")
  obs <- subj |>
    tidyr::crossing(time = sort(unique(c(
      seq(0, tau * n_doses, by = 3),
      seq(tau * 55, tau * 56, by = 0.1)   # Day-56 interval, sampled finely
    )))) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
  dplyr::bind_rows(dosing, obs) |> dplyr::arrange(id, time, dplyr::desc(evid))
}))

sim_wt <- rxode2::rxSolve(mod_typ, events = wt_events,
                          keep = c("treatment", "WT")) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc))

sim_wt |>
  dplyr::mutate(treatment = factor(treatment, levels = paste0(wt_arms, " kg"))) |>
  dplyr::filter(time <= tau * n_doses) |>
  ggplot(aes(time / 24, Cc, colour = treatment)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = 56, linetype = "dashed", colour = "grey40") +
  labs(x = "Time (days)", y = "Deutenzalutamide (ug/mL)", colour = "Body weight",
       title = "Figure 4 - 80 mg once daily at different body weights",
       caption = paste("Replicates Figure 4 of Liu 2024. Dashed line: Day 56,",
                       "the day Liu 2024 reports steady state was reached."))
```

# PKNCA validation

## Steady-state and first-dose NCA by dose group

```{r pknca-cohort}
sim_nca <- sim_obs |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time = 0 record per subject; pre-dose Cc = 0 is correct for an
# extravascular model. Existing time-0 rows win via .keep_all on the first row.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

dose_df <- events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "ug/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "mg")

intervals <- data.frame(
  start   = c(0, t_last_dose),
  end     = c(tau, t_last_dose + tau),
  cmax    = TRUE,
  tmax    = TRUE,
  auclast = TRUE,
  ctrough = TRUE
)

nca_res <- suppressWarnings(
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
)

nca_tbl <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "ctrough")) |>
  dplyr::mutate(period = ifelse(start == 0, "First dose", "Day 84 (steady state)"))

nca_summary <- nca_tbl |>
  dplyr::group_by(period, treatment, PPTESTCD) |>
  dplyr::summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median)

nca_summary |>
  dplyr::mutate(treatment = factor(treatment, levels = paste0(doses_mg, " mg"))) |>
  dplyr::arrange(period, treatment) |>
  dplyr::rename(
    "Period"                 = period,
    "Dose"                   = treatment,
    "AUC0-tau (ug*h/mL)"     = auclast,
    "Cmax (ug/mL)"           = cmax,
    "Ctrough (ug/mL)"        = ctrough,
    "Tmax (h)"               = tmax
  ) |>
  knitr::kable(digits = 2,
               caption = "Median simulated NCA by dose group, first dose and Day 84.")
```

Liu 2024 (Exploratory Data Analysis) reports that "the average peak time of
multiple administration was about 1 h (0.5-2 hours)" and that "the peak time was
not affected by the change of dose level".

```{r tmax-check}
tmax_ss <- nca_summary |>
  dplyr::filter(period == "Day 84 (steady state)") |>
  dplyr::pull(tmax)
stopifnot(length(tmax_ss) == length(doses_mg))
# The paper's own stated range, used as an absolute bound; not a bound taken
# from one run. Realised medians sit near 1.05-1.15 h across arms.
stopifnot(all(tmax_ss >= 0.5), all(tmax_ss <= 2))
```

Liu 2024 (Discussion) reports that "the plasma concentration of deutenzalutamide
after stabilization was 16.5 times that of a single dose".

```{r accumulation}
acc <- nca_tbl |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(id, treatment, period, PPORRES) |>
  tidyr::pivot_wider(names_from = period, values_from = PPORRES) |>
  dplyr::mutate(Rac = `Day 84 (steady state)` / `First dose`)

acc |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(`Median accumulation ratio (AUC)` = median(Rac), .groups = "drop") |>
  dplyr::mutate(treatment = factor(treatment, levels = paste0(doses_mg, " mg"))) |>
  dplyr::arrange(treatment) |>
  dplyr::rename("Dose" = treatment) |>
  knitr::kable(digits = 1,
               caption = "Accumulation ratio vs. Liu 2024's reported 16.5-fold.")

rac_med <- median(acc$Rac)
# Liu 2024's 16.5 is quoted for "plasma concentration", not specifically for
# AUC0-tau; the AUC, Cmax and Ctrough accumulation ratios of this model differ
# from one another (about 17, 9 and 19 respectively), so a generous band is
# used deliberately. It still fails on a mis-transcribed CL/F or V1/F, either
# of which moves the ratio by a factor.
stopifnot(rac_med > 10, rac_med < 25)
```

## Comparison against Liu 2024's published simulation

Liu 2024 (Results, Simulation) simulates 80 mg once daily in virtual subjects
weighing 50, 60, 70 and 80 kg and reports the steady-state AUC over the dosing
interval as 476, 458, 426 and 414 ug*h/mL. Liu 2024 states that steady state
was reached on Day 56, so the Day-56 dosing interval is used here; the paper
does not name the day it evaluated.

```{r pknca-weight}
wt_nca <- sim_wt |>
  dplyr::filter(time >= tau * 55, time <= tau * 56) |>
  dplyr::select(id, time, Cc, treatment)

wt_dose <- wt_events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, treatment)

wt_conc_obj <- PKNCA::PKNCAconc(wt_nca, Cc ~ time | treatment + id,
                                concu = "ug/mL", timeu = "h")
wt_dose_obj <- PKNCA::PKNCAdose(wt_dose, amt ~ time | treatment + id,
                                doseu = "mg")

wt_intervals <- data.frame(
  start   = tau * 55,
  end     = tau * 56,
  auclast = TRUE,
  cmax    = TRUE,
  tmax    = TRUE
)

wt_res <- suppressWarnings(
  PKNCA::pk.nca(PKNCA::PKNCAdata(wt_conc_obj, wt_dose_obj, intervals = wt_intervals))
)

published <- tibble::tribble(
  ~treatment, ~auclast, ~tmax,
  "50 kg",    476,      1.0,
  "60 kg",    458,      1.0,
  "70 kg",    426,      1.0,
  "80 kg",    414,      1.0
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = wt_res,
  reference     = published,
  by            = "treatment",
  units         = c(auclast = "ug*h/mL", tmax = "h"),
  tolerance_pct = 20
)

knitr::kable(cmp, digits = 1,
             caption = paste("Simulated vs. Liu 2024 Results (Simulation) and",
                             "Exploratory Data Analysis. * differs by >20%."),
             align = c("l", "l", "r", "r", "r"))
```

```{r pknca-weight-gate}
sim_auc <- as.data.frame(wt_res$result) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::arrange(match(treatment, published$treatment)) |>
  dplyr::pull(PPORRES)
stopifnot(length(sim_auc) == 4L)

pct <- 100 * (sim_auc / published$auclast - 1)
names(pct) <- published$treatment
print(round(pct, 1))

# These are typical-value (zeroRe) solves: fully deterministic, so no cohort
# noise enters and a tight bound is appropriate. The residual gap is the
# unstated evaluation day -- Liu 2024's numbers sit between the Day-52 and
# Day-56 intervals. A mis-transcribed CL/F, dose, or body-weight reference
# moves these by tens of percent.
stopifnot(max(abs(pct)) < 8)
```

Liu 2024 further reports that across 50-80 kg "the fluctuation range of
steady-state exposure ... was 96.2%-110.6%" relative to the typical body
weight, i.e. a highest-to-lowest ratio of 1.150.

```{r fluctuation}
spread_sim <- max(sim_auc) / min(sim_auc)
spread_pub <- max(published$auclast) / min(published$auclast)
tibble::tibble(
  Source = c("Liu 2024 Discussion", "This model (Day 56)"),
  `Highest / lowest AUC0-tau across 50-80 kg` = c(spread_pub, spread_sim)
) |>
  knitr::kable(digits = 3)

stopifnot(abs(spread_sim / spread_pub - 1) < 0.10)
```

## Internal identity: AUC0-tau at true steady state

For a linear model, `AUC0-tau` at true steady state equals `Dose / (CL/F)`
exactly, independent of every distribution parameter. This is the sharpest
available check on `CL/F` and on the dose-to-concentration scaling.

```{r ss-identity}
ss_events <- tibble::tibble(id = 1L, WT = 64) |>
  tidyr::crossing(time = tau * seq(0, 299)) |>
  dplyr::mutate(amt = 80, evid = 1L, cmt = "depot") |>
  dplyr::bind_rows(
    tibble::tibble(id = 1L, WT = 64) |>
      tidyr::crossing(time = seq(tau * 299, tau * 300, by = 0.05)) |>
      dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
  ) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

ss_sim <- rxode2::rxSolve(mod_typ, events = ss_events) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(tad = time - tau * 299)

auc_ss    <- auc_tau(ss_sim)
auc_ident <- 80 / cl_typ

tibble::tibble(
  Quantity = c("Simulated AUC0-tau after 300 daily doses (ug*h/mL)",
               "Dose / (CL/F) (ug*h/mL)"),
  Value = c(auc_ss, auc_ident)
) |>
  knitr::kable(digits = 1)

# Deterministic; the only error is trapezoidal discretisation of a sharp peak.
stopifnot(abs(auc_ss / auc_ident - 1) < 0.02)
```

# Known deviation: the reported V2/F covariate range

Liu 2024 (Results, Simulation) states that, relative to the typical 64 kg,
`V2/F` was "34.9% lower" at the 5% body-weight quantile and "102% higher" at the
95% quantile. Neither statement is reproducible from Table 1 together with the
Table S3 body-weight range, and this is recorded as a deviation rather than
gated.

```{r v2-deviation}
e_wt <- theta[["e_wt_vp"]]
wt_grid <- c(46, 50, 64, 81)
tibble::tibble(
  `Body weight (kg)` = wt_grid,
  `V2/F (L)`         = vp_typ * (wt_grid / 64)^e_wt,
  `% vs 64 kg`       = 100 * ((wt_grid / 64)^e_wt - 1)
) |>
  knitr::kable(digits = 1,
               caption = paste("V2/F across the full Table S3 body-weight range",
                               "(46-81 kg) using Table 1's exponent of 1.83."))
```

At the *maximum* observed weight of 81 kg the model gives `V2/F` only 53.9%
above the 64 kg typical value; reaching "+102%" would require a body weight of
about 94 kg, which is 13 kg above the cohort maximum. The "-34.9%" figure
corresponds to about 50.6 kg and is consistent with a 5% quantile of the
observed 46-81 kg range, so the low end is reproducible and the high end is
not. Either the "102%" figure is a typographical error, or it combines the
covariate effect with the `V2/F` between-subject variability; the model file
encodes Table 1 and Equation 3 as printed and does not attempt to reconcile the
sentence.

# Assumptions and deviations

- **`omega` and `sigma` in Table 1 are variances, not standard deviations.**
  Liu 2024's Statistical Model text says "omega represents the standard
  deviation", but the tabulated numbers are the NONMEM `$OMEGA` / `$SIGMA`
  *variances*. Read as standard deviations they would imply 2.7% CV on `CL/F`
  and a 2.1% proportional residual error - the latter below the assay's own
  reported imprecision (intra-assay 0.6-4.7%, inter-assay 3.4-4.9%, Liu 2024
  Bioanalytical Methods), which is impossible for a residual term that also
  absorbs model misspecification. Read as variances they give 16.7%, 21.3% and
  74.3% CV and a 14.4% proportional error, all plausible. The model file encodes
  them as variances.
- **Equation 3 is a power model, not an exponential one.** Liu 2024's Covariate
  Model section calls Equation 3 "exponential mode", but the printed equation is
  `theta_i = theta_TV * (COV_i / COV_bar)^theta_2`, a ratio-exponent (power)
  form. The printed equation governs. In practice the two forms are numerically
  almost indistinguishable over 46-81 kg (they differ by under 1% in `AUC0-tau`
  at every weight tested), so this choice does not drive any result here.
- **Body-weight reference of 64.0 kg.** Liu 2024 does not print `COV_bar` in
  Equation 3. It is taken from the Results (Simulation) sentence naming "the
  typical value of body weight (64.0 kg)", which coincides with the Table S3
  median.
- **Virtual body-weight distribution.** Table S3 reports only the median (64 kg)
  and range (46-81 kg). A normal distribution with SD 9 kg truncated to that
  range is assumed for the cohort; nothing in the validation depends on the
  shape beyond covering the observed range.
- **Evaluation day for the body-weight simulation.** Liu 2024 reports
  steady-state `AUC0-tau` at four body weights but does not say which day was
  simulated. The Day-56 interval is used because the paper states steady state
  was reached on Day 56. The published values actually sit closest to the Day-52
  interval; using Day 56 is the pre-specified rather than the best-fitting
  choice.
- **Table 1 unit labels are inconsistent with the quantities.** `CL/F` and
  `CLd/F` are labelled "(L*h)" but are clearances (L/h); `ka` is labelled
  "(hour)" but is a rate constant (1/h); and the covariate row "BW on V2/F (L)"
  is a dimensionless exponent. These are table-formatting artefacts; the model
  file uses the dimensionally correct units.
- **Table S3's AST row is internally inconsistent** - it reports a range of
  13-18 U/L with a median of 27 U/L. AST is not in the final model, so this does
  not affect the extraction; it is recorded in `covariatesDataExcluded` and
  noted here for completeness.
- **`V1/F` and `CLd/F` are fixed.** Table 1 marks both "(fixed)" and reports no
  RSE, bootstrap median or bootstrap CI for either. Liu 2024 does not state
  where the fixed values came from. They are encoded with `fixed()`.
- **The metabolites M1 and M2 are not modelled.** They were measured
  bioanalytically (Liu 2024 Bioanalytical Methods, Figure S1) but Liu 2024
  develops a parent-only population PK model.
- **Screened-but-not-retained covariates** (age, height, serum creatinine,
  leukocytes, neutrophils, platelets, total bilirubin, AST, ALT, haemoglobin,
  ECOG status) are recorded in the model file's `covariatesDataExcluded` list
  with their Table S4 objective-function changes, and are deliberately absent
  from `covariateData`.
- **Extrapolation limits.** Liu 2024's Discussion states that all subjects had
  renal and hepatic laboratory values within the normal range, so the model
  should not be used to predict exposure in hepatic or renal impairment.
- Every parameter value comes from Liu 2024's own text, Table 1, or the
  supplement; none was digitised from a figure or supplied by correspondence.
