---
title: "Nifedipine (Liu 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nifedipine (Liu 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Liu_2022_nifedipine"))$reference`
* Description: `r rxode2::rxode(readModelDb("Liu_2022_nifedipine"))$description`
* Article: <https://doi.org/10.3389/fphar.2022.970539> (open access, PMC9462537)

Apatinib, a VEGFR-2 tyrosine kinase inhibitor, causes hypertension in most
treated patients, and nifedipine is the antihypertensive most often used for
it. Apatinib also inhibits CYP3A4, the main clearance pathway of nifedipine,
and raises nifedipine exposure. Liu 2022 asks whether that pharmacokinetic
interaction also produces a clinically important *pharmacodynamic*
interaction -- an excessive fall in blood pressure -- by chaining three pieces:

1. a Simcyp v16 minimal-PBPK model of nifedipine (immediate-release tablet with
   first-order absorption; controlled-release tablet with the ADAM absorption
   model and a Weibull dissolution profile);
2. the authors' previously published Simcyp PBPK model of apatinib (Liu 2021),
   linked to nifedipine through competitive CYP3A4 inhibition
   (`Ki = 0.12 uM`, `fu,mic = 0.65`);
3. an **ordinary Emax model** (equation 3) relating the nifedipine plasma
   concentration to the change in systolic blood pressure (SBP).

**Only the third component, the PD layer, is packaged in nlmixr2lib.** The
reasoning is in [Assumptions and deviations](#assumptions-and-deviations): the
two PBPK layers are Simcyp platform models whose clearance is entered per pmol
of recombinant enzyme and whose controlled-release absorption depends on a
dissolution profile the paper does not print, so they cannot be reproduced from
the publication. The PD layer is fully specified by the paper and is
self-contained once the nifedipine plasma concentration is supplied through the
`CEFFECT` covariate.

```{r model}
mod <- readModelDb("Liu_2022_nifedipine")
mod
```

# Population

No subject-level data were fitted. Every simulation in the paper used Simcyp
virtual populations of 100 subjects aged 26-65 years, half of them female (24
subjects of the Sim-Chinese population for the DDI verification against Zhu
2020). The PD layer was checked against published mean SBP changes in
hypertensive patients after a single 60 mg controlled-release (CR) tablet and a
single 10 mg immediate-release (IR) tablet (Figure 3, Table 5), and was then
applied to virtual cancer patients (Table 6, Figure 4) and to patients with
Child-Pugh A, B and C hepatic impairment (Table 7), each with and without
apatinib 750 mg once daily.

```{r population}
str(rxode2::rxode(readModelDb("Liu_2022_nifedipine"))$population)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Liu_2022_nifedipine.R`. The table
below collects them in one place.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `dsbp <- -emax * conc / (ec50 + conc)` | n/a | Equation 3, Methods "Development and verification of pharmacodynamic model for nifedipine" |
| `conc <- CEFFECT` (direct link, no effect compartment) | n/a | Equation 3 uses the concentration `C` directly; Discussion: "a close and direct relationship between circulating drug concentrations and antihypertensive effect" |
| `lemax` (Emax) | 30 mmHg magnitude (paper: -30 mmHg), fixed | Methods PD paragraph "set at -30 mmHg"; Results "When the Emax was set to -30 mmHg, the PD model fitted best" |
| `lec50` (EC50) | 12.12 ng/mL, fixed | Methods PD paragraph "set at ... 12.12 ng/ml"; Discussion "the EC50 value was set to 12.12 ng/ml in this study" |
| `addSd` | 0 (fixed) | No residual-error model is reported |
| (no IIV) | n/a | No inter-individual variability is reported for the PD parameters |

## Units

| Symbol | Quantity | Units |
|--------|----------|-------|
| `CEFFECT` / `conc` | nifedipine plasma concentration | ng/mL (EC50 is given in ng/ml) |
| `ec50` | half-maximal plasma concentration | ng/mL |
| `emax`, `dsbp` | reduction / change in systolic blood pressure | mmHg (Table 5 and Figure 3 units) |
| `time` | time | h (Figures 3 and 4) |

`conc / (ec50 + conc)` is a ratio of concentrations and therefore unitless, so
`dsbp` carries the units of `emax` (mmHg). The model has no ODE state and no
dose record; `time` only orders the records and does not enter any equation.

# Simulation

The model is an algebraic map from a nifedipine plasma concentration to a
change in SBP. A simulation is therefore a set of observation records carrying
the `CEFFECT` covariate. Because the model declares no `eta`, do **not** pass
`omega = NA` or apply `rxode2::zeroRe()`.

```{r simulate}
conc_values <- seq(0, 250, by = 0.5)
conc_grid <- data.frame(
  id = 1L,
  time = seq_along(conc_values),
  CEFFECT = conc_values,
  evid = 0L,
  amt = 0
)

sim <- rxode2::rxSolve(mod, events = conc_grid) |>
  as.data.frame()

head(sim[, c("CEFFECT", "dsbp")])
```

## Concentration-effect curve

The paper does not plot the concentration-effect curve itself, so the figure
below marks on it the mean predicted nifedipine Cmax values the paper prints for
its PBPK simulations (Tables 3 and 4).

```{r ce-curve, fig.width = 6.5, fig.height = 4.2}
printed_cmax <- tibble::tribble(
  ~scenario, ~CEFFECT,
  "60 mg CR, healthy (Table 3)", 26.51,
  "20 mg IR, healthy (Table 3)", 144.59,
  "30 mg CR alone (Table 4)", 20.97,
  "30 mg CR + apatinib (Table 4)", 29.51
) |>
  mutate(dsbp = -30 * CEFFECT / (12.12 + CEFFECT))

ggplot(sim, aes(CEFFECT, dsbp)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = -30, linetype = "dashed", colour = "grey50") +
  geom_point(data = printed_cmax, aes(colour = scenario), size = 2.6) +
  labs(
    x = "Nifedipine plasma concentration (ng/mL)",
    y = "Change in SBP (mmHg)",
    colour = NULL,
    title = "Liu 2022 equation 3: change in SBP versus nifedipine concentration",
    caption = paste(
      "Dashed line: Emax = -30 mmHg.",
      "Points: mean predicted Cmax printed in Liu 2022 Tables 3 and 4."
    )
  ) +
  theme(legend.position = "bottom", legend.direction = "vertical")
```

# Validation

NCA is not an applicable validation for this model: there is no dose, no
compartment and no concentration-time profile to integrate. The checks below
follow the pattern for mechanistic models: definitional identities, boundary
behaviour, and agreement with the summary values the paper prints. Every
quantity is **deterministic** (the model has no random effects), so the
identity checks use tight tolerances; where a bound is looser it is because the
paper's value comes from a different calculation, explained alongside.

## Definitional identities

These hold by algebra and must match to solver precision. A wrong `ec50` moves
the half-maximal point and breaks the second identity immediately; a wrong
`emax` breaks the second and third.

```{r definitional}
anchor <- rxode2::rxSolve(
  mod,
  events = data.frame(
    id = 1L, time = 1:3, evid = 0L, amt = 0,
    CEFFECT = c(0, 12.12, 1e9)
  )
) |>
  as.data.frame()

identities <- tibble::tibble(
  Property = c(
    "C = 0 gives no change",
    "C = EC50 gives Emax / 2",
    "C -> infinity gives Emax"
  ),
  Expected = c(0, -15, -30),
  Model = anchor$dsbp
)
knitr::kable(identities, digits = 6,
             caption = "Definitional identities of the ordinary Emax model.")

stopifnot(
  isTRUE(all.equal(anchor$dsbp[1], 0)),
  isTRUE(all.equal(anchor$dsbp[2], -15, tolerance = 1e-10)),
  isTRUE(all.equal(anchor$dsbp[3], -30, tolerance = 1e-6))
)
```

The change in SBP must also fall monotonically with concentration and can never
go beyond Emax:

```{r shape}
stopifnot(
  all(diff(sim$dsbp) < 0),
  all(sim$dsbp > -30)
)
```

## Maximum SBP reduction against the paper's printed exposures

With a direct Emax link the largest SBP reduction a subject experiences,
`Rmax`, occurs at that subject's Cmax, so `Rmax_i = f(Cmax_i)`. The paper
prints mean predicted Cmax for the 30 mg CR DDI design (Table 4) and mean
predicted Rmax for 30 mg CR in virtual cancer patients with and without
apatinib (Table 6). Because `f` is concave, the mean of the individual `Rmax`
values is *smaller in magnitude* than `f(mean Cmax)` (Jensen's inequality), so
the model evaluated at the printed mean Cmax should slightly overshoot the
printed mean Rmax, never undershoot it.

```{r rmax-ddi}
ddi <- tibble::tibble(
  Arm = c("30 mg CR alone", "30 mg CR + apatinib"),
  `Mean Cmax, Table 4 (ng/mL)` = c(20.97, 29.51),
  `Mean Rmax, Table 6 (mmHg)` = c(-18.62, -20.30)
)

ddi_sim <- rxode2::rxSolve(
  mod,
  events = data.frame(
    id = 1L, time = 1:2, evid = 0L, amt = 0,
    CEFFECT = ddi$`Mean Cmax, Table 4 (ng/mL)`
  )
) |>
  as.data.frame()

ddi <- ddi |>
  mutate(
    `Model f(mean Cmax) (mmHg)` = ddi_sim$dsbp,
    `Difference (%)` = 100 * (`Model f(mean Cmax) (mmHg)` -
      `Mean Rmax, Table 6 (mmHg)`) / `Mean Rmax, Table 6 (mmHg)`
  )
knitr::kable(ddi, digits = 2,
             caption = "Model SBP reduction at the printed mean Cmax versus the printed mean Rmax.")

stopifnot(
  # Jensen direction: the model at the mean Cmax is at least as large a
  # reduction as the paper's mean of individual maxima.
  all(ddi$`Model f(mean Cmax) (mmHg)` <= ddi$`Mean Rmax, Table 6 (mmHg)`),
  # Size: a wrong EC50 or Emax moves these by tens of percent. Achieved:
  # +2.1% and +4.8%.
  all(abs(ddi$`Difference (%)`) < 10)
)
```

Table 4 comes from the Sim-Chinese DDI-verification population (24 subjects)
and Table 6 from the virtual cancer-patient population (100 subjects), so the
two tables do not describe exactly the same cohort; the agreement within 5% is
nevertheless what the paper's Emax and EC50 imply.

## The pharmacodynamic interaction is attenuated by saturation

The paper's central conclusion is that a 1.41-fold rise in nifedipine Cmax
(and 1.73-fold rise in AUC) with apatinib changes the SBP response by no more
than 1.25-fold. Equation 3 alone explains why: for a concentration increase by
a factor R, the effect increases by the factor R (EC50 + C) / (EC50 + R C), which
is R at very low concentration and falls towards 1 as `C` approaches and
passes EC50.

```{r attenuation}
ratio_model <- ddi_sim$dsbp[2] / ddi_sim$dsbp[1]
attenuation <- tibble::tibble(
  Quantity = c(
    "Cmax ratio, with / without apatinib (Table 4)",
    "Rmax ratio, paper (Table 6, 30 mg CR)",
    "Rmax ratio, model at the Table 4 mean Cmax"
  ),
  Value = c(1.41, 1.09, ratio_model)
)
knitr::kable(attenuation, digits = 3)

stopifnot(
  isTRUE(all.equal(
    ratio_model,
    1.41 * (12.12 + 20.97) / (12.12 + 1.41 * 20.97),
    tolerance = 1e-2
  )),
  ratio_model < 1.25,
  abs(ratio_model - 1.09) < 0.05
)
```

```{r attenuation-plot, fig.width = 6.5, fig.height = 4}
base_conc <- seq(0.5, 150, by = 0.5)
atten_curve <- bind_rows(
  tibble::tibble(CEFFECT = base_conc, r = 1.41),
  tibble::tibble(CEFFECT = base_conc, r = 1.73)
) |>
  mutate(
    effect_ratio = r * (12.12 + CEFFECT) / (12.12 + r * CEFFECT),
    label = paste0("Exposure ratio ", r)
  )

ggplot(atten_curve, aes(CEFFECT, effect_ratio, colour = label)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 1.25, linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 12.12, linetype = "dotted", colour = "grey50") +
  labs(
    x = "Nifedipine concentration without apatinib (ng/mL)",
    y = "Ratio of SBP change, with / without apatinib",
    colour = NULL,
    caption = paste(
      "Dashed line: the paper's 1.25-fold criterion.",
      "Dotted line: EC50 = 12.12 ng/mL."
    )
  )
```

## Comparison with the PD-verification table (Table 5)

Table 5 compares predicted and observed Rmax in hypertensive patients. The
paper does not print the nifedipine Cmax of the hypertensive-patient
simulations, so the nearest printed exposures are the healthy-volunteer
predictions of Table 3. For the 10 mg IR tablet, the 20 mg IR Cmax is halved
(the PBPK model has linear clearance at these doses). These are indicative
comparisons across populations, not validation gates.

```{r table5}
t5 <- tibble::tibble(
  `Dose (Table 5)` = c("60 mg CR", "10 mg IR"),
  `Exposure used (ng/mL)` = c(26.51, 144.59 / 2),
  `Exposure source` = c("Table 3, 60 mg CR Cmax", "Table 3, 20 mg IR Cmax / 2"),
  `Predicted Rmax, Table 5 (mmHg)` = c(-22.21, -30.13),
  `Observed Rmax, Table 5 (mmHg)` = c(-23.09, -33.20)
) |>
  mutate(`Model f(Cmax) (mmHg)` = -30 * `Exposure used (ng/mL)` /
    (12.12 + `Exposure used (ng/mL)`))
knitr::kable(t5, digits = 2)
```

The printed predicted Rmax of -30.13 mmHg for 10 mg IR is *beyond* the model's
Emax of -30 mmHg, and the 5th percentile of the predicted curve in Figure 3B
reaches about -33 mmHg. No concentration can produce that with a single
deterministic Emax of -30 mmHg, so the paper's Simcyp PD simulation must have
carried between-subject variability on Emax and/or EC50 that the paper does not
report. This is recorded under
[Assumptions and deviations](#assumptions-and-deviations).

# Assumptions and deviations

* **PBPK layers not reproduced.** The nifedipine PK layer is a Simcyp v16
  minimal-PBPK model. Its clearance is entered as recombinant CYP3A4 and CYP3A5
  Vmax (pmol/min/pmol of enzyme) and Km (Table 1). Turning those into a hepatic
  and gut-wall clearance needs the enzyme abundances, microsomal protein per
  gram of liver, liver weight, hepatic blood flow and the intersystem
  extrapolation factors of the Simcyp population library, none of which the
  paper prints, and no in-vivo clearance is given to anchor them. The CR tablet
  is absorbed through the ADAM model with a Weibull dissolution profile taken
  from Doki 2017 whose parameters are not printed either. The apatinib layer is
  a separate Simcyp model (Liu 2021) whose clearance is also entered per mg of
  microsomal protein. The PK layers therefore cannot be reproduced from the
  publication, and the nifedipine plasma concentration is supplied through
  `CEFFECT`. The printed Table 1 compound inputs are kept in the `CEFFECT` notes
  of the model file for provenance.
* **Total plasma concentration.** Equation 3 uses "nifedipine concentration"
  without saying total or unbound. Every concentration the paper reports is
  total plasma (Tables 3 and 4, Figures 1 and 2), and the EC50 of 12.12 ng/mL
  is a total-plasma literature value, so `CEFFECT` is total plasma.
* **Emax and EC50 are fixed.** Both were set rather than estimated: EC50 from
  the literature, Emax chosen as the best of a range of literature values by
  comparing predicted with observed SBP changes. Neither carries a standard
  error, so both are wrapped in `fixed()`.
* **No between-subject variability.** The paper reports none for the PD layer,
  but its predicted mean Rmax for 10 mg IR (-30.13 mmHg, Table 5) is beyond
  Emax, and Figures 3 and 4 show 5th-95th percentile bands, so the Simcyp PD
  simulation must have included variability on Emax and/or EC50 (probably the
  platform default). Its magnitude is not reported and is not invented here;
  the model is typical-value only and `addSd` is fixed at zero.
* **Sign convention.** The paper's Emax is signed (-30 mmHg). The model
  carries the magnitude on the log scale (`lemax = log(30)`) and applies the
  sign in `model()`, so `dsbp` is negative for a reduction, matching
  equation 3 and Figures 3 and 4.
* **AUE is not checked.** The area under the effect-time curve (Tables 5 to 7)
  depends on the full predicted concentration-time course of the virtual
  populations, which the paper shows only as raster mean curves for a different
  population (Figure 2, Sim-Chinese DDI design) than its AUE tables. The
  windows over which AUE was integrated are also not stated (Tables 5 and 6
  give 433.98 and 787.55 mmHg*h for the same 60 mg CR Rmax of -22.21 mmHg).
* **No erratum.** A Europe PMC search on 2026-10-09 found no correction notice
  for this article and no supplementary material.
