---
title: "REC-2282 (AR-42) (Liva 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{REC-2282 (AR-42) (Liva 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Liva_2021_rec2282"))$reference`
* Description: `r rxode2::rxode(readModelDb("Liva_2021_rec2282"))$description`
* Article: <https://doi.org/10.1007/s13318-021-00722-z> (open access; PMC8599380)

REC-2282 (development code AR-42, formerly OSU-HDAC42) is an oral
pan-histone-deacetylase inhibitor with no international non-proprietary name, so
the model is named for its current development code.

# Population

Liva 2021 pooled two phase 1 trials run at The Ohio State University
Comprehensive Cancer Center: the first-in-human trial OSU09102 (NCT01129193;
multiple myeloma, lymphoma and advanced solid tumours including
neurofibromatosis type 2) and OSU11130 (NCT01798901; relapsed or refractory acute
myeloid leukaemia). Of 57 enrolled patients (44 + 13; Table 1), 882 plasma
concentrations from 55 were modelled (Results 3.2). Table 2 summarises the 56
patients with anthropometric data: median weight 76.1 kg (42.9-122.4), median
fat-free mass 50.8 kg (29.9-76.0), median BMI 26.0 kg/m^2, median age 63 years
(20-80), 51.8% female. Supplemental Table 2 lists 40 patients with a haematologic
malignancy (multiple myeloma 17, lymphoma 10, AML 13) and 16 with a solid tumour.
Doses of 20-80 mg were taken fasted three times weekly (Monday, Wednesday,
Friday; one OSU11130 cohort 40 mg four times weekly) for 3 weeks of each 28-day
cycle, with intensive sampling on day 1 and on day 5 (OSU11130) or day 19
(OSU09102).

The same information is available programmatically via
`readModelDb("Liva_2021_rec2282")()$population`.

# Source trace

Every `ini()` value carries an in-file comment pointing to its source; the table
collects them.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Structure: depot -(ka)-> transit -(ka)-> central <-> peripheral, lag on the dose | n/a | Fig. 2; Results 3.2; Discussion |
| `lcl` (CL/F) | log(11.6) L/h | Table 3, covariate model |
| `lvc` (Vc/F) | log(105) L | Table 3, covariate model |
| `lq` (Q/F) | log(7.1) L/h | Table 3, covariate model |
| `lvp` (Vp/F) | log(76.5) L | Table 3, covariate model |
| `lka` (ka) | log(1.28) 1/h | Table 3, covariate model |
| `ltlag` (ALAG) | log(0.0453) h | Table 3, covariate model (no IIV in the final model, Results 3.2) |
| `e_ffm_cl` | 0.493 | Table 3 'Covariates FFM'; power form of Equation 1 |
| `e_heme_tlag` | 0.942 | Table 3 'Covariates TMR'; linear form of Equation 2 |
| `e_tablet_tlag` | 0.74 | Table 3 'Covariates FORM'; linear form of Equation 2 |
| FFM centring value | 50.8 kg | Table 2 median FFM, all patients; Section 2.3 (median normalisation) |
| FFM equation | Janmahasatian | Section 2.4 |
| `etalcl` | 0.05068 | Table 3: 22.8 CV%, as log(CV^2 + 1) |
| `etalka` | 0.20500 | Table 3: 47.7 CV%, as log(CV^2 + 1) |
| `expSd` | 0.545 | Table 3 epsilon (SD, per footnote); Section 2.3 'additive error model for log-transformed data' |

# Virtual cohort

The observed data are not public. The virtual cohort reproduces the Table 2
demographics: body weight log-normal around the median of 76.1 kg, BMI normal
with the Table 2 mean and SD, 51.8% female, and fat-free mass computed from
these by the Janmahasatian equations the paper uses (Section 2.4). Tumour type
follows the Supplemental Table 2 split (40 of 56 haematologic). The paper does not
report how many patients took each formulation, so half the cohort is assigned
capsules.

Four dose levels are simulated on the paper's three-times-weekly schedule
(doses on days 1, 3, 5, 8, 10, 12, 15, 17 and 19), with 100 virtual patients per
arm.

```{r cohort}
set.seed(20211007)
dose_levels <- c(20, 40, 50, 60)
n_per_arm <- 100L

ffm_janmahasatian <- function(wt, bmi, sexf) {
  ifelse(
    sexf == 1,
    9270 * wt / (8780 + 244 * bmi),
    9270 * wt / (6680 + 216 * bmi)
  )
}

dose_days <- c(1, 3, 5, 8, 10, 12, 15, 17, 19)
dose_times <- (dose_days - 1) * 24
obs_rel <- c(0, 0.25, 0.5, 1, 1.5, 2, 3, 4, 6, 8, 10, 12, 16, 24)
obs_times <- sort(unique(c(
  obs_rel, 36, 47.99,          # day 1 (to the day 3 dose)
  96 + obs_rel,                # day 5
  432 + obs_rel, 432 + 36, 480 # day 19
)))

make_cohort <- function(n, dose, id_offset) {
  subj <- tibble(
    id = id_offset + seq_len(n),
    SEXF = rbinom(n, 1, 0.518),
    WT = pmin(pmax(exp(rnorm(n, log(76.1), 0.23)), 42.9), 122.4),
    BMI = pmin(pmax(rnorm(n, 26.7, 5.4), 18.5), 43.6),
    TUMTP_SOLID = 1L - rbinom(n, 1, 40 / 56),
    FORM_CAPSULE = rbinom(n, 1, 0.5),
    treatment = paste(dose, "mg")
  ) |>
    mutate(FFM = ffm_janmahasatian(WT, BMI, SEXF))
  doses <- tidyr::crossing(subj, time = dose_times) |>
    mutate(evid = 1L, amt = dose, cmt = "depot")
  obs <- tidyr::crossing(subj, time = obs_times) |>
    mutate(evid = 0L, amt = 0, cmt = "central")
  bind_rows(doses, obs) |>
    arrange(id, time, desc(evid))
}

events <- bind_rows(lapply(seq_along(dose_levels), function(i) {
  make_cohort(n_per_arm, dose_levels[i], id_offset = (i - 1L) * n_per_arm)
}))
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))

cohort <- distinct(events, id, treatment, SEXF, WT, BMI, FFM, TUMTP_SOLID)
summary(cohort$FFM)
```

The simulated median FFM (`r round(median(cohort$FFM), 1)` kg) lies close to the
Table 2 median of 50.8 kg, so the virtual cohort is centred where the CL/F
covariate term is anchored.

# Simulation

```{r simulate}
mod <- readModelDb("Liva_2021_rec2282")
sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep = c("treatment", "FFM", "TUMTP_SOLID", "FORM_CAPSULE")
) |>
  as.data.frame() |>
  mutate(
    day = case_when(
      time < 48 ~ "Day 1",
      time >= 96 & time <= 120 ~ "Day 5",
      time >= 432 ~ "Day 19",
      TRUE ~ NA_character_
    ),
    tad = case_when(
      day == "Day 1" ~ time,
      day == "Day 5" ~ time - 96,
      day == "Day 19" ~ time - 432
    )
  )
```

`Cc` is the individual prediction without residual error.

# Replicate published figures

```{r figure-1, fig.width = 7, fig.height = 7}
# Replicates Figure 1 of Liva 2021: mean concentration by dose level on days 1, 5 and 19.
sim |>
  filter(!is.na(day)) |>
  group_by(day, treatment, time) |>
  summarise(Cc = mean(Cc), .groups = "drop") |>
  mutate(day = factor(day, c("Day 1", "Day 5", "Day 19"))) |>
  ggplot(aes(time, Cc, colour = treatment)) +
  geom_line() +
  facet_wrap(~day, ncol = 1, scales = "free_x") +
  scale_y_log10(limits = c(1, 1000)) +
  labs(
    x = "Time after first dose (h)", y = "REC-2282 (ng/mL)", colour = "Dose",
    caption = "Replicates Figure 1 of Liva 2021 (lines = mean of each dose group)."
  )
```

Figure 1 of the paper shows day-1 mean peaks of roughly 100 ng/mL at 20 mg to
300-400 ng/mL at 50-70 mg, falling to about 3-10 ng/mL by 48 h; the simulated
means follow the same pattern.

```{r figure-3, fig.width = 7, fig.height = 4}
# Replicates the style of Figure 3 of Liva 2021 (VPC) for the 40 mg arm, which
# contributed the most patients (Table 1).
set.seed(1)
sim |>
  filter(treatment == "40 mg", !is.na(day)) |>
  mutate(
    obs = Cc * exp(rnorm(n(), 0, 0.545)),
    day = factor(day, c("Day 1", "Day 5", "Day 19"))
  ) |>
  group_by(day, tad) |>
  summarise(
    Q05 = quantile(obs, 0.05), Q50 = quantile(obs, 0.50),
    Q95 = quantile(obs, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~day) +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)", y = "REC-2282 (ng/mL)",
    caption = "Simulated 5th/50th/95th percentiles with residual error, 40 mg arm (cf. Figure 3 of Liva 2021)."
  )
```

## Supplemental Figure 3: typical AUC at a 40 mg flat dose

Supplemental Figure 3 shows simulated AUCs for the patients dosed at 40 mg flat.
The box is narrow (whiskers about 2.7-3.2 h*mg/L, median about 2.85 h*mg/L), which
indicates typical-value predictions that vary only by each patient's FFM. The
paper does not state the AUC window, so the typical-value AUC over 0-24 h, 0-48 h
(the dosing interval) and 0-infinity is shown for the cohort median FFM.

```{r supp-fig-3}
mod_typ <- rxode2::zeroRe(mod)
ev_typ <- rxode2::et(amt = 40, cmt = "depot") |>
  rxode2::et(seq(0, 480, by = 0.05), cmt = "central") |>
  as.data.frame() |>
  mutate(FFM = 50.8, TUMTP_SOLID = 1, FORM_CAPSULE = 1)
typ <- as.data.frame(rxode2::rxSolve(mod_typ, ev_typ))
trap <- function(t, y) sum(diff(t) * (head(y, -1) + tail(y, -1)) / 2)
auc_window <- function(end) {
  x <- typ[typ$time <= end, ]
  trap(x$time, x$Cc) / 1000
}
typ_auc <- tibble(
  window = c("0-24 h", "0-48 h", "0-infinity (dose / CL)"),
  auc = c(auc_window(24), auc_window(48), 40 / 11.6)
)
knitr::kable(
  typ_auc |> rename("AUC window" = window, "Typical AUC (h*mg/L)" = auc),
  digits = 2,
  caption = "Typical-value AUC after 40 mg at FFM = 50.8 kg; Supplemental Figure 3 median is about 2.85 h*mg/L."
)
# Structural gate: the digitised median (2.85) must lie between the shortest
# and the longest plausible window. A mis-transcribed CL or unit moves both
# bounds by the same factor and breaks this.
stopifnot(typ_auc$auc[1] < 2.85 * 1.1, typ_auc$auc[3] > 2.85 * 0.9)
```

The digitised median of 2.85 h*mg/L falls between the typical AUC0-24 and the
AUC0-infinity. (The trapezoid sum above is a check on a single noiseless
profile; the cohort NCA below uses PKNCA.)

# PKNCA validation

NCA is run on the day-1 and day-19 dosing days over 0-24 h after the dose, per
dose level.

```{r pknca}
sim_nca <- sim |>
  filter(!is.na(Cc), day %in% c("Day 1", "Day 19"), tad <= 24) |>
  select(id, time, Cc, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)
dose_df <- events |>
  filter(evid == 1) |>
  select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = c(0, 432),
  end = c(24, 456),
  cmax = TRUE,
  tmax = TRUE,
  auclast = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_df <- as.data.frame(nca_res$result) |>
  mutate(day = ifelse(start == 0, "Day 1", "Day 19"))

nca_df |>
  filter(PPTESTCD %in% c("cmax", "tmax", "auclast")) |>
  group_by(treatment, day, PPTESTCD) |>
  summarise(median = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  rename(
    "Dose" = treatment, "Day" = day, "Cmax (ng/mL)" = cmax,
    "Tmax (h)" = tmax, "AUC0-24 (ng*h/mL)" = auclast
  ) |>
  knitr::kable(digits = 1, caption = "Simulated median NCA by dose level and day.")
```

## Comparison against published NCA

Liva 2021 does not tabulate NCA. The first-in-human report of the haematologic
arm of OSU09102 (Sborov et al., Leuk Lymphoma 2017;58:2310-2318, doi:10.1080/10428194.2017.1298751) gives day-1 and
day-19 geometric-mean Cmax of 0.383 and 0.335 uM at 20 mg, 0.794 and 0.830 uM at
40 mg, and 1.60 and 1.50 uM at 50 mg. These are converted to ng/mL with the
REC-2282 molecular weight of 312.4 g/mol (C18H20N2O3). The simulated median is
compared with the geometric mean, which is close to the median for log-normal
data.

```{r nca-compare}
mw <- 312.37
published <- tibble::tribble(
  ~treatment, ~day, ~cmax,
  "20 mg", "Day 1", 0.383,
  "20 mg", "Day 19", 0.335,
  "40 mg", "Day 1", 0.794,
  "40 mg", "Day 19", 0.830,
  "50 mg", "Day 1", 1.60,
  "50 mg", "Day 19", 1.50
) |>
  mutate(cmax = cmax * mw)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_df |> filter(treatment %in% published$treatment),
  reference = published,
  by = c("treatment", "day"),
  params = "cmax",
  units = c(cmax = "ng/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Simulated vs. published Cmax (Sborov 2017). * differs by >20%.")
```

The 40 mg arm agrees within 5% on both days, as does day 1 at 20 mg. The flagged
20 mg day-19 row reflects a published day-19 value below the day-1 value in a
three-patient arm; the linear model predicts slight accumulation instead. The 50 mg arm of Sborov 2017 had
geometric-mean Cmax values about double those at 40 mg, which is more than dose
proportional. Liva 2021 reports dose-proportional PK across 20-80 mg in the
pooled data (Discussion), so the model, which is linear, falls short of that
small arm by roughly the dose ratio. The 40 mg arm has the most patients and is
used as the gate:

```{r nca-gate}
cmax40 <- nca_df |>
  filter(treatment == "40 mg", PPTESTCD == "cmax")
# Centre gate: a mis-transcribed Vc, ka or dose unit shifts the median Cmax by
# tens of percent. Robust to which subjects land in the tails.
stopifnot(abs(median(cmax40$PPORRES) / (0.794 * mw) - 1) < 0.2)

# Variability gate. The Discussion reports that observed dose-normalised AUC
# varied by 24.1%. The simulated between-subject CV of AUC0-24 comes from IIV on
# CL/F (22.8% CV) and ka plus the FFM effect. It sits somewhat lower (about
# 15-17% per arm when this vignette was written) because a 24 h window is less
# sensitive to CL/F than AUC0-infinity and the observed value also carries assay
# noise. The bounds are an envelope, not a match, and were checked to go red:
# with the IIV removed the CV falls to about 8%, and reading the Table 3 CV% as
# a variance (omega^2 = 0.228, 0.477) raises it to about 28%.
auc_d1 <- nca_df |> filter(PPTESTCD == "auclast", day == "Day 1")
cv_auc <- auc_d1 |>
  group_by(treatment) |>
  summarise(cv = sd(PPORRES) / mean(PPORRES) * 100, .groups = "drop")
cv_auc
stopifnot(median(cv_auc$cv) > 11, median(cv_auc$cv) < 22)
```

## Covariate effects on the lag time

```{r lag-table}
tibble(
  group = c(
    "Solid tumour, capsule", "Solid tumour, tablet",
    "Haematologic, capsule", "Haematologic, tablet"
  ),
  TUMTP_SOLID = c(1, 1, 0, 0),
  FORM_CAPSULE = c(1, 0, 1, 0)
) |>
  mutate(alag_min = 60 * 0.0453 * (1 + 0.942 * (1 - TUMTP_SOLID)) * (1 + 0.74 * (1 - FORM_CAPSULE))) |>
  select(group, alag_min) |>
  rename("Group" = group, "Typical lag time (min)" = alag_min) |>
  knitr::kable(digits = 1)
```

The lag time is 3-9 minutes in every group, which is short against the
absorption half-life of about 30 minutes (ln 2 / 1.28 1/h, through two
compartments). The tumour-type and formulation effects therefore barely change
the simulated profiles.

# Assumptions and deviations

- **Coding of the tumour-type and formulation covariates.** Liva 2021 evaluates
  tumour type 'as dichotomous variables for solid vs. heme tumor' and defines
  FORM as 'formulation (capsule vs. tablet)', but never says which level is
  coded 1. It has no reference-subject statement, no control stream, and no supplement
  table that settles it. The maintainers took the first-named level as the
  reference (0): solid tumour for tumour type and capsule for formulation. So
  haematologic malignancy and the tablet carry the fractional increases of
  0.942 and 0.74 on ALAG. The model expresses both through the existing
  canonical indicators, whose value 1 marks the reference level here:
  `ALAG = 0.0453 * (1 + 0.942 * (1 - TUMTP_SOLID)) * (1 + 0.74 * (1 - FORM_CAPSULE))`.
  With the opposite coding, the
  typical lag would sit with haematologic patients on tablets instead. The
  practical effect is small because every group's typical lag is under 10
  minutes.
- **Formulation mix.** The paper does not report which patients took capsules
  or tablets; the virtual cohort assigns 50% to each.
- **IIV scale.** Table 3 reports IIV as CV%. The variances use
  `omega^2 = log(CV^2 + 1)`, which is appropriate for the exponential IIV model in
  Section 2.3. Reading the CV as `sqrt(omega^2)` instead would change the
  variances by 4% (CL) and 10% (ka).
- **Residual error.** Table 3 labels epsilon 'proportional', while Section 2.3
  describes an 'additive error model for log-transformed data'. The two are the
  same model on the linear scale, encoded here as a log-normal residual with
  `expSd = 0.545` (an SD, per the Table 3 footnote).
- **Lag placement.** The Discussion says the lag was added 'into the transit
  compartment', while Fig. 2 draws it on the oral dose entering the depot. For a
  linear chain the two are equivalent (the input is simply delayed); the model
  applies `alag(depot)`.
- **FFM centring.** Section 2.3 normalises continuous covariates by the
  population median. The Table 2 median over the 56 patients with anthropometrics
  (50.8 kg) is used; the modelled population was 55 patients.
- **Non-paper-derived values in this vignette.** The external Cmax comparison
  uses Sborov et al. 2017 (the OSU09102 haematologic cohort) and the REC-2282
  molecular weight of 312.4 g/mol (C18H20N2O3). Neither enters the model.
- No erratum or correction notice for Liva 2021 was found on Crossref or
  Europe PMC (checked 2026-09-29).
