---
title: "Rozanolixizumab anti-FcRn PK/PD, monkey to human (Lledo-Garcia 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rozanolixizumab anti-FcRn PK/PD, monkey to human (Lledo-Garcia 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
rxode2::rxSetSeed(20220112)
```

# Model and source

Rozanolixizumab is a humanized IgG4P monoclonal antibody against the human
neonatal Fc receptor (FcRn). By occupying FcRn it blocks the salvage pathway
that normally recycles endogenous IgG, so circulating IgG (including pathogenic
autoantibodies) is diverted to lysosomal degradation. Lledo-Garcia 2022 builds a
semi-mechanistic PK/PD model of that system in cynomolgus monkey, translates it
to human by allometry to design the first-in-human (FIH) study, and then
re-estimates it against the FIH data.

The paper therefore reports **three distinct parameter sets for one structure**,
and this package ships each as its own model file (see
`references/replicate-author-structure.md` -- the author's structure is
reproduced rather than collapsed):

| Model | What it is | Source |
|---|---|---|
| `LledoGarcia_2022_rozanolixizumab_cyno` | Fitted to single-dose IV cynomolgus monkey data | Table 1, first value column |
| `LledoGarcia_2022_rozanolixizumab_translated` | Allometric forward projection to a 75 kg human; used to design the FIH study, before any human data existed | Table 1, second value column |
| `LledoGarcia_2022_rozanolixizumab` | Final model, re-estimated against the FIH IV cohorts | Table 2, FIH-estimated column |

```{r models}
mods <- lapply(
  c(cyno       = "LledoGarcia_2022_rozanolixizumab_cyno",
    translated = "LledoGarcia_2022_rozanolixizumab_translated",
    final      = "LledoGarcia_2022_rozanolixizumab"),
  function(nm) rxode2::rxode(readModelDb(nm))
)
```

* Citation: `r mods$final$reference`
* Article: <https://doi.org/10.1002/psp4.12739>
* Supplement (Texts S1-S4, Figures S1-S2): <https://doi.org/10.1002/psp4.12739>, "Supporting Information"

# Population

The **cynomolgus** model was fitted to single-dose IV data at 5, 10 and 30 mg/kg.
The underlying animal study is Smith et al. 2018 (mAbs 10:1111-1130; reference 9
of the source paper), which used four healthy young adult male animals per
dosing group weighing at least 2.3 kg and aged roughly 135-185 weeks, so the
three single-dose IV groups contribute 12 animals. Multiple-dose data collected
before antidrug antibodies were detected were held back for external validation.

The **final human** model was fitted to the IV cohorts of the FIH study
NCT02220153 (Kiessling et al. 2017, Sci Transl Med 9:eaan1208; reference 11).
That study randomised 49 healthy subjects to rozanolixizumab (n = 36) or placebo
(n = 13) across six cohorts; the first three were IV at 1, 4 and 7 mg/kg with
n = 6 each, and Lledo-Garcia 2022 analysed only the IV cohorts. Dose escalation
was stopped at 7 mg/kg on tolerability grounds, so the model sees a narrower
dose range than planned. Lledo-Garcia 2022 does not tabulate baseline
demographics for this cohort.

The same information is available programmatically from each model's
`population` metadata.

```{r population}
str(mods$final$population, max.level = 1)
```

# Model structure

The structure is identical across all three parameter sets: a two-compartment
target-mediated disposition (TMDD) model under the quasi-equilibrium
approximation, coupled to an indirect-response model in which **free** drug
stimulates IgG catabolism. Figure 1 of the paper is the schematic.

Two structural simplifications are load-bearing (Methods, and Supplementary
Text S1):

1. FcRn is highly expressed on vascular endothelium, so the endosomal
   compartment is lumped with the central compartment and binding is assumed to
   equilibrate instantly.
2. Rozanolixizumab binds FcRn through its Fab, and the clearance of free FcRn
   equals the clearance of the drug:FcRn complex (both `CLB`). Because those two
   are equal, the **total** FcRn pool is constant in time, so the paper carries
   no differential equation for it -- `TB` enters the binding quadratic as a
   constant.

Supplementary Text S1 and S2 give the equations (recovered from the MathType
objects embedded in the supplement's Word file):

```
dTA/dt    = -FA(t)*CLA/V - CPLX(t)*CLB/V - Q*FA(t)/V + Q*APERI(t)/V2 + input
dAPERI/dt =  Q*FA(t)/V - Q*APERI(t)/V2

CPLX  = ( (KD*V + TA(t) + TB) - sqrt( (KD*V + TA(t) + TB)^2 - 4*TA(t)*TB ) ) / 2
FA    = TA - CPLX
CLB   = KDEG*V ;  RB = CFCRN*KDEG*V ;  TB = V*RB/CLB

EFF     = EMAX*FAC(t) / (FAC(t) + EC50)      with FAC = MWT*FA/V
dIGG/dt = IGSS*KOUT - KOUT*IGG(t)*(1 + EFF)
```

The packaged model divides the amount-scale binding quadratic through by `V`,
which is an algebraic identity (the expression is homogeneous of degree one), so
that drug and target are carried on a single mg/L (= ug/mL) scale with dosing in
mg. The free-drug root is written in the numerically stable arrangement
`cfree = (disc + sqrt(disc^2 + 4*kd*ctot))/2`.

```{r structure}
mods$final$funTxt
```

# Source trace

Per-parameter provenance is recorded as an in-file comment beside each `ini()`
entry. Collected here for review:

| Quantity | cyno | translated | final (FIH) | Source |
|---|---|---|---|---|
| `V` (L) | 0.108 | 2.55 | 2.7 | Table 1 / Table 2 |
| `Q` (L/day) | 0.0221 | 0.237 | 0.271 | Table 1 / Table 2 |
| `V2` (L) | 0.0162 | 0.383 | 0.36 | Table 1 / Table 2 |
| `CLA` (L/day) | 0.0709 | 0.7605 | 0.968 | Table 1 / Table 2 |
| `KD` (nmol/L) | 0.827 | 0.359 | 1 | Table 1 / Table 2 |
| `[FcRn]` (nmol/L) | 592.89 | 592.89 | 147 | Table 1 / Table 2 |
| `Kdeg` (1/day) | 0.3389 | 0.277 | 0.88 | Table 1 / Table 2 |
| `IgG base` (g/L) | 10.5 | 11.65 | 9.88 | Table 1 / Table 2 |
| `Kout` (1/day) | 0.0431 | 0.031 | 0.0364 | Table 1 / Table 2 |
| `Emax` | 7.58 | 7.58 | 4.24 | Table 1 / Table 2 |
| `EC50` (mg/L) | 1.14 | 0.4959 | 0.154 | Table 1 / Table 2 |
| IIV (CV%) | not reported | V 16.35, V2 28.75, CLA 35.15, Kdeg 18, IgG base 21, Kout 22, Emax 25, EC50 58 | V 15.8, Kdeg 30.3, IgG base 18.4 | Table 1 / Table 2 |
| Proportional RUV | not reported | not reported | PK 9.3%, IgG 5.8% | Table 2 |
| PK / PD equations | Supplementary Text S1 / S2 | | | supplement |
| Allometry | Supplementary Text S4 | | | supplement |
| `MWT` = 148 kDa | **not in the paper** -- RYSTIGGO US prescribing information section 11 | | | see Errata |

# Gate 1: the derived-quantity identities

Table 1 and Table 2 print `CLB` and `RB` as *derived* quantities
(Table 2 footnote b: "Derived value (not estimated)"), with
`RB = [FcRn]*Kdeg*V` and `CLB = Kdeg*V`. The packaged models compute both from
their primaries, so reproducing the printed numbers is an exact, deterministic
check on the volume, the target concentration and the degradation rate all at
once -- no seed, no cohort, no solver tolerance involved.

```{r gate-derived}
one_obs <- function(mod, dose_mg) {
  ev <- rbind(
    data.frame(id = 1L, time = 0, amt = dose_mg, evid = 1L,
               cmt = "central", dvid = NA_integer_),
    data.frame(id = 1L, time = c(0, 1), amt = NA_real_, evid = 0L,
               cmt = "central", dvid = 1L)
  )
  as.data.frame(rxode2::rxSolve(rxode2::zeroRe(mod), ev, returnType = "data.frame"))
}

derived <- tibble::tibble(
  Model     = c("cyno", "translated", "final"),
  RB_model  = vapply(mods, function(m) one_obs(m, 1)$rb[1], numeric(1)),
  RB_paper  = c(21.7, 419.748, 349.3)
) |>
  mutate(pct_diff = 100 * (RB_model - RB_paper) / RB_paper)

derived |>
  rename("RB, model (nmol/day)" = RB_model,
         "RB, paper (nmol/day)" = RB_paper,
         "Difference (%)"       = pct_diff) |>
  knitr::kable(digits = c(0, 2, 3, 3),
               caption = "RB = [FcRn] x Kdeg x V reproduced from the packaged primaries.")

# The translated set is the loosest because the paper rounds Kdeg to 0.277
# (the value implied by the printed RB and CLB is 0.2776).
stopifnot(all(abs(derived$pct_diff) < 0.5))
```

# Gate 2: the translated model reproduces Figure 3a

Figure 3a overlays the translated model's predictions on the observed FIH
concentrations, and its y axis is labelled **"Free drug concentration
(ug/mL)"**. That label is what settles an otherwise ambiguous modelling choice:
the PK observation is the *free* drug concentration, not total. It is also a
sharp quantitative test, because free and total differ by orders of magnitude at
the low dose where FcRn is far from saturated.

```{r gate-figure3a}
peak_free <- function(mod, dose_mg) {
  s <- one_obs(mod, dose_mg)
  c(free = s$Cc[1], total = s$Ctot[1])
}

fig3a <- tibble::tibble(dose_mgkg = c(1, 4, 7)) |>
  rowwise() |>
  mutate(pk = list(peak_free(mods$translated, dose_mgkg * 75))) |>
  ungroup() |>
  mutate(free_ugmL  = vapply(pk, function(x) x[["free"]],  numeric(1)),
         total_ugmL = vapply(pk, function(x) x[["total"]], numeric(1))) |>
  select(-pk)

fig3a |>
  rename("Dose (mg/kg)"          = dose_mgkg,
         "Peak FREE (ug/mL)"     = free_ugmL,
         "Peak TOTAL (ug/mL)"    = total_ugmL) |>
  knitr::kable(digits = c(0, 4, 1),
               caption = "Translated model, peak concentration after a single IV dose.")
```

Reading the model medians off Figure 3a gives roughly 0.02, 20-30 and
150 ug/mL at 1, 4 and 7 mg/kg. The free-drug column matches all three; the
total-drug column is wrong by a factor of about 1100 at 1 mg/kg and about 4 at
4 mg/kg. This is the check that pins the observation variable.

```{r gate-figure3a-assert}
stopifnot(
  # 1 mg/kg: deeply target-bound, free drug is a small fraction of total.
  fig3a$free_ugmL[1] > 0.01, fig3a$free_ugmL[1] < 0.05,
  # 4 mg/kg: FcRn just saturating.
  fig3a$free_ugmL[2] > 20,   fig3a$free_ugmL[2] < 40,
  # 7 mg/kg: well past saturation.
  fig3a$free_ugmL[3] > 100,  fig3a$free_ugmL[3] < 150,
  # The total-drug reading is excluded by the same figure.
  fig3a$total_ugmL[1] > 25
)
```

# Gate 3: the final model reproduces the observed IgG reductions

The Results section reports the FIH outcome directly: "mean reductions in IgG of
14.5, 33.4, and 47.6% (10 days postdose) were observed for the 1, 4, and 7 mg/kg
rozanolixizumab doses". The prediction-corrected VPCs of Figure 4 confirm the
final model described these data. Because these are the paper's own headline
numbers and the final model was fitted to them, reproducing them from the
packaged typical-value parameters exercises the whole chain: the binding
quadratic, the unit conversion through the molecular weight, and the
indirect-response PD.

```{r gate-igg10}
igg_profile <- function(mod, dose_mg, tmax = 84) {
  tt <- sort(unique(c(seq(0, 2, by = 0.05), seq(2, 14, by = 0.25),
                      seq(14, tmax, by = 1))))
  ev <- rbind(
    data.frame(id = 1L, time = 0, amt = dose_mg, evid = 1L,
               cmt = "central", dvid = NA_integer_),
    data.frame(id = 1L, time = tt, amt = NA_real_, evid = 0L,
               cmt = "central", dvid = 1L)
  )
  as.data.frame(rxode2::rxSolve(rxode2::zeroRe(mod), ev, returnType = "data.frame"))
}

day10 <- lapply(c(1, 4, 7), function(d) {
  s <- igg_profile(mods$final, d * 75, tmax = 20)
  tibble::tibble(dose_mgkg = d,
                 model_pct = -s$pctChangeIgG[which.min(abs(s$time - 10))])
}) |>
  bind_rows() |>
  mutate(observed_pct = c(14.5, 33.4, 47.6),
         diff_pp      = model_pct - observed_pct)

day10 |>
  rename("Dose (mg/kg)"                 = dose_mgkg,
         "Model IgG reduction (%)"      = model_pct,
         "Observed IgG reduction (%)"   = observed_pct,
         "Difference (percentage pts)"  = diff_pp) |>
  knitr::kable(digits = c(0, 1, 1, 1),
               caption = "Day-10 IgG reduction, final model typical value vs the observed FIH means (Results).")

stopifnot(
  # Every dose within 5 percentage points of the observed mean.
  all(abs(day10$diff_pp) < 5),
  # And the dose-response ordering is preserved.
  !is.unsorted(day10$model_pct)
)
```

# Gate 4: the preclinical model spans the observed monkey response

The Discussion states that the monkey data "demonstrated 49-75% decreases in IgG
from baseline (following the rozanolixizumab dose range studied)" of 5, 10 and
30 mg/kg. Converting mg/kg to mg needs an animal body weight, which the paper
does not report; the weight implied by its own allometric translation
(`V` 0.108 L to 2.55 L at exponent 1, and `V2` 0.0162 L to 0.383 L, both a
23.6-fold ratio against 75 kg) is about 3.17 kg, and that is what is used here.

The gated claims are therefore the ones that are robust to that unknown: the
top of the dose range reaches the paper's upper figure, and the response is
monotone in dose. The lower bound is *not* gated -- see Errata.

```{r gate-cyno}
bw_cyno <- 75 / (2.55 / 0.108)   # kg, implied by the paper's own allometry

cyno_nadir <- lapply(c(5, 10, 30), function(d) {
  s <- igg_profile(mods$cyno, d * bw_cyno, tmax = 60)
  tibble::tibble(dose_mgkg = d, nadir_pct = -min(s$pctChangeIgG))
}) |>
  bind_rows()

cyno_nadir |>
  rename("Dose (mg/kg)" = dose_mgkg, "Maximum IgG reduction (%)" = nadir_pct) |>
  knitr::kable(digits = 1,
               caption = paste0("Preclinical model, maximum IgG reduction at an implied ",
                                round(bw_cyno, 2), " kg body weight."))

stopifnot(
  # 30 mg/kg reaches the top of the observed 49-75% band.
  cyno_nadir$nadir_pct[3] > 65, cyno_nadir$nadir_pct[3] < 85,
  # Monotone dose-response, which no body-weight assumption can disturb.
  !is.unsorted(cyno_nadir$nadir_pct)
)
```

# Virtual cohort and stochastic simulation

Original observed data are not available. The VPC-style figures below use
virtual cohorts of 100 subjects per dose arm, matching the FIH IV dose levels.
A 75 kg reference subject is used throughout because the models carry no body
weight covariate -- allometry was applied once, when translating the monkey
parameters, and is not part of the model.

```{r cohort}
make_arm <- function(dose_mgkg, n = 100, id_offset = 0L, tmax = 84) {
  tt <- sort(unique(c(seq(0, 2, by = 0.1), seq(2, 14, by = 0.5),
                      seq(14, tmax, by = 2))))
  ids <- id_offset + seq_len(n)
  dose <- data.frame(id = ids, time = 0, amt = dose_mgkg * 75, evid = 1L,
                     cmt = "central", dvid = NA_integer_)
  obs <- expand.grid(id = ids, time = tt)
  obs <- data.frame(id = obs$id, time = obs$time, amt = NA_real_, evid = 0L,
                    cmt = "central", dvid = 1L)
  out <- rbind(dose, obs)
  out$treatment <- paste0(dose_mgkg, " mg/kg IV")
  out[order(out$id, out$time, -out$evid), ]
}

doses <- c(1, 4, 7)
events <- bind_rows(lapply(seq_along(doses), function(i) {
  make_arm(doses[i], n = 100, id_offset = (i - 1L) * 1000L)
}))
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

`rxSolve` on an `rxUi` scales super-linearly with the number of subjects in a
single call, so each arm is solved separately and the results are stacked.

```{r simulate}
solve_arm <- function(mod, ev) {
  as.data.frame(
    rxode2::rxSolve(mod, events = ev, keep = c("treatment"),
                    returnType = "data.frame")
  )
}

sim_final <- bind_rows(lapply(split(events, events$treatment), function(ev) {
  solve_arm(mods$final, ev)
}))
sim_translated <- bind_rows(lapply(split(events, events$treatment), function(ev) {
  solve_arm(mods$translated, ev)
}))
```

# Replicate published figures

```{r figure-3a-vpc}
# Replicates Figure 3a of Lledo-Garcia 2022: translated-model free drug
# concentration vs time, by IV dose cohort.
sim_translated |>
  filter(time > 0, Cc > 0) |>
  group_by(treatment, time) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = median(Cc),
            Q95 = quantile(Cc, 0.95), .groups = "drop") |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue") +
  facet_wrap(~treatment, scales = "free_y") +
  scale_y_log10() +
  coord_cartesian(xlim = c(0, 6)) +
  labs(x = "Time (days)", y = "Free drug concentration (ug/mL)",
       title = "Figure 3a - translated model, free rozanolixizumab",
       caption = "Replicates Figure 3a of Lledo-Garcia 2022. Median with 90% prediction interval.")
```

```{r figure-3b-vpc}
# Replicates Figure 3b of Lledo-Garcia 2022: percentage change from baseline IgG.
sim_translated |>
  filter(time > 0) |>
  group_by(treatment, time) |>
  summarise(Q05 = quantile(pctChangeIgG, 0.05), Q50 = median(pctChangeIgG),
            Q95 = quantile(pctChangeIgG, 0.95), .groups = "drop") |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "darkgreen") +
  geom_line(colour = "darkgreen") +
  geom_hline(yintercept = 0, linetype = "dotted") +
  facet_wrap(~treatment) +
  labs(x = "Time (days)", y = "Change from baseline IgG (%)",
       title = "Figure 3b - translated model, IgG response",
       caption = "Replicates Figure 3b of Lledo-Garcia 2022. Median with 90% prediction interval.")
```

```{r figure-4}
# Replicates Figure 4 of Lledo-Garcia 2022: final-model prediction-corrected VPC
# equivalent, for rozanolixizumab concentration (a) and IgG concentration (b).
bind_rows(
  sim_final |> transmute(treatment, time, value = Cc,
                         panel = "(a) Free rozanolixizumab (ug/mL)"),
  sim_final |> transmute(treatment, time, value = IgG_obs,
                         panel = "(b) Total IgG (g/L)")
) |>
  filter(time > 0, value > 0) |>
  group_by(panel, treatment, time) |>
  summarise(Q025 = quantile(value, 0.025), Q50 = median(value),
            Q975 = quantile(value, 0.975), .groups = "drop") |>
  ggplot(aes(time, Q50, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = Q025, ymax = Q975), alpha = 0.2, colour = NA) +
  geom_line() +
  facet_wrap(~panel, scales = "free_y") +
  labs(x = "Time (days)", y = NULL, colour = NULL, fill = NULL,
       title = "Figure 4 - final human model",
       caption = paste("Replicates Figure 4 of Lledo-Garcia 2022.",
                       "Median with 2.5th-97.5th percentile band."))
```

# PKNCA validation

The paper's central pharmacological claim is that the system is **nonlinear at
low dose and moves towards linearity as dose increases** ("Prediction/observation
alignment increased at higher doses, as the system moved towards linearity for
both PK and IgG"; the FIH report likewise describes "nonlinear increases with
dose"). That is a directly testable NCA statement: because FcRn acts as a
saturable sink for free drug, **dose-normalised** free-drug exposure must
*increase* with dose, and the increase must flatten out once the receptor is
saturated.

The NCA window is 0-14 days. Free drug is essentially gone well before that --
at 1 mg/kg the median free concentration has fallen from about 6.8 ug/mL at
time zero to about 1e-6 ug/mL by day 14, so roughly 23 half-lives -- and beyond
about day 28 the profile decays into solver noise and drifts marginally
negative (order 1e-12 ug/mL). Truncating keeps the assertion below meaningful
rather than silently repairing negative values, and the omitted tail
contributes nothing to AUC.

```{r pknca}
sim_nca <- sim_final |>
  filter(!is.na(Cc), time <= 14) |>
  select(id, time, Cc, treatment)

# Guarantee a time-zero row per (id, treatment) so PKNCA can anchor AUC.
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

# TMDD tails can drift marginally negative in the solver; PKNCA would take
# log() of those. Assert instead of silently repairing.
stopifnot(all(sim_nca$Cc >= 0))

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- events |>
  filter(evid == 1) |>
  select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(start = 0, end = 14,
                        cmax = TRUE, tmax = TRUE, auclast = TRUE)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
```

```{r pknca-summary}
nca_tab <- as.data.frame(nca_res) |>
  group_by(treatment, PPTESTCD) |>
  summarise(median = median(PPORRES), .groups = "drop") |>
  pivot_wider(names_from = PPTESTCD, values_from = median) |>
  mutate(dose_mg = as.numeric(sub(" mg/kg IV", "", treatment)) * 75,
         auc_per_mg  = auclast / dose_mg,
         cmax_per_mg = cmax / dose_mg) |>
  arrange(dose_mg)

nca_tab |>
  select(treatment, cmax, tmax, auclast, auc_per_mg, cmax_per_mg) |>
  rename("Dose cohort"                    = treatment,
         "Cmax (ug/mL)"                   = cmax,
         "Tmax (day)"                     = tmax,
         "AUC0-14 (ug*day/mL)"            = auclast,
         "AUC0-14 / dose (ug*day/mL/mg)"  = auc_per_mg,
         "Cmax / dose (ug/mL/mg)"         = cmax_per_mg) |>
  knitr::kable(digits = c(0, 2, 2, 1, 4, 5),
               caption = paste("Median NCA of simulated FREE rozanolixizumab,",
                               "final model, 100 subjects per arm."))
```

```{r pknca-assert}
# Dose-normalised exposure must rise with dose: that IS the TMDD nonlinearity.
stopifnot(
  !is.unsorted(nca_tab$auc_per_mg),
  !is.unsorted(nca_tab$cmax_per_mg),
  # The nonlinearity is large between 1 and 4 mg/kg ...
  nca_tab$auc_per_mg[2] / nca_tab$auc_per_mg[1] > 2,
  # ... and much smaller between 4 and 7 mg/kg, i.e. approaching linearity.
  nca_tab$auc_per_mg[3] / nca_tab$auc_per_mg[2] <
    nca_tab$auc_per_mg[2] / nca_tab$auc_per_mg[1]
)
```

## Comparison against published NCA

Lledo-Garcia 2022 reports **no** NCA table -- neither Cmax, Tmax, AUC nor
half-life is tabulated for either species, so there is nothing to place
side by side with the simulated values above. The quantitative comparisons
against published numbers in this vignette are therefore Gates 1-4, which use
the paper's derived-parameter identities (Table 1, Table 2 footnote b), its
Figure 3a concentrations and its Results-section IgG reductions instead.

# Assumptions and deviations

## Errata and inconsistencies found in the source

* **Table 2 misprints the translated `Kdeg` as 0.227 /day.** Table 1 gives
  0.277 /day, and both derived quantities printed alongside it confirm Table 1:
  `CLB = Kdeg*V` requires 0.7079 / 2.55 = 0.2776, and `RB = [FcRn]*Kdeg*V`
  requires 419.748 / (592.89 x 2.55) = 0.2776. The packaged translated model
  uses 0.277.
* **Table 1 and Table 2 disagree on the translated `Q`** (0.237 vs 0.27 L/day).
  Table 1's value is the allometrically consistent one
  (0.0221 x (2.55/0.108)^0.75 = 0.2367) and Table 1 is the table the paper
  describes as the input to the simulations, so 0.237 is used.
* **Figure 3's panel labels are transposed between the caption and the figure.**
  The caption says "(a) percentage change from baseline in IgG ... and (b)
  rozanolixizumab concentration"; the rendered figure has concentration in (a)
  and IgG in (b). This vignette follows the figure.
* **Supplementary Text S2 gives `IGSS` in mg/L** while Tables 1 and 2 print IgG
  baseline in mg/mL (10.5, 11.65, 9.88 -- physiologically the mg/mL reading is
  the correct one for serum IgG). The IgG differential equation is homogeneous
  of degree one in IgG, so the choice of scale cannot affect any prediction; the
  models carry IgG in g/L (= mg/mL) as printed in the tables.

## Non-paper-derived parameter values

* **`MWT` = 148 kDa (rozanolixizumab molecular weight).** This value is *not
  reported anywhere on disk for this paper*: not in the main text, not in
  Supplementary Texts S1-S4, and not in the cited characterisation paper
  (Smith et al. 2018, mAbs 10:1111-1130), which was retrieved and searched. It
  is nonetheless load-bearing, because the paper reports `KD` and `[FcRn]` in
  nmol/L but `EC50` and all observed concentrations in mg/L, so exactly one
  molar-to-mass conversion is needed to put drug and target on a common scale
  in the binding quadratic. The value used is from the RYSTIGGO
  (rozanolixizumab-noli) US prescribing information, section 11 DESCRIPTION:
  "Rozanolixizumab-noli has an approximate molecular weight of 148 kDa"
  (DailyMed SPL setid 3c0eb8c2-c042-4954-b451-3baa77f5e6d1). It is hardcoded in
  `model()` as a unit conversion rather than declared in `ini()`, following the
  `Said_2025_imatinib` precedent for molecular-weight conversion factors.
  Gate 2 above is an independent check on it: the free concentrations the
  packaged model produces at 1, 4 and 7 mg/kg agree with Figure 3a, which they
  would not if the conversion were materially wrong.

## Modelling assumptions

* **The PK observation is free, not total, drug.** The paper never states this
  in prose. It is established here by the y-axis label of Figure 3a ("Free drug
  concentration (ug/mL)"), corroborated by Gate 2, and consistent with the
  bioanalytical method: Smith et al. 2018 describe an assay in which
  rozanolixizumab is captured on immobilised FcRn, i.e. one that reports
  FcRn-binding-competent drug. Total drug is still available from the packaged
  models as the derived variable `Ctot`.
* **Residual error is encoded as `fixed(0)` in the cynomolgus and translated
  models.** Table 2 prints "-" for both proportional-RUV rows of the translated
  column, and no RUV magnitude is reported for the monkey fit; the translated
  model was used for simulation rather than fitted. No value was invented.
* **The cynomolgus model carries no IIV.** Table 1's "IIV human (CV %)" column
  describes the human translation, not the monkey fit, and no monkey IIV is
  reported anywhere. The model is therefore a typical-value mechanism.
* **The cynomolgus body weight is not reported.** Only "a minimum of 2.3 kg"
  appears in Smith et al. 2018. Gate 4 uses 3.17 kg, the weight implied by the
  paper's own allometric translation. The predicted IgG nadir at 5 mg/kg is
  sensitive to this choice (39% at 3.17 kg, 50% at 4.5 kg) whereas the 30 mg/kg
  nadir is not (74% to 77% across the same range). Gate 4 therefore asserts on
  the 30 mg/kg endpoint and on dose-monotonicity, and deliberately does not
  assert the lower edge of the paper's observed 49-75% band. The body weight was
  **not** tuned to make that band fit.
* **Q, V2, KD and [FcRn] in the final model are estimates, not fixed values.**
  Table 2 footnote a marks them as "estimated with prior information" (NONMEM
  NWPRIOR) and each carries an RSE, so none is wrapped in `fixed()`.
* **`units$dosing` is mg and `units$concentration` is ug/mL.** The conventions
  checker notes the magnitude difference; it is correct here because an amount
  in mg divided by a volume in L gives mg/L, which is numerically identical to
  ug/mL. No scaling factor is needed or applied.
* **The 1 mg/kg dose is known to be underpredicted.** The paper says so
  explicitly for both the translated and the final model, and attributes it to
  the estimated FcRn concentration having greater leverage at low dose, with no
  monkey data below 5 mg/kg to inform the nonlinear region. This is a property
  of the published model, faithfully reproduced, not an extraction defect.
