---
title: "Vancomycin (Ma 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vancomycin (Ma 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Ma_2020_vancomycin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Ma_2020_vancomycin"))$description`
* Article: <https://doi.org/10.3389/fphar.2020.563967> (open access, CC BY 4.0)

# Population

Ma 2020 is a retrospective, single-centre therapeutic-drug-monitoring (TDM)
study at the First Affiliated Hospital of Zhejiang University School of
Medicine, Hangzhou. Fifty-six adult kidney transplant recipients (35 male, 21
female; all grafts from brain-dead donors) received intravenous vancomycin as
postoperative prophylaxis after surgery between March and June 2017 (Methods,
"Patients and Data Collection"). Table 1 reports age 43.72 +/- 9.92 years
(range 24-70), body weight 58.27 +/- 8.47 kg (median 59.95, range 37.7-79),
serum creatinine median 164 umol/L (range 60-1,490) and estimated GFR median
39.91 mL/min (IQR 32.53-59.44, range 3.38-108.61). Nearly all patients received
prednisone, mycophenolate mofetil and tacrolimus (four received cyclosporine A
instead).

Vancomycin was given as 500 mg per administration: one dose on the first
postoperative day, then 1-4 administrations per day (500-2,000 mg/day) adjusted
by TDM (Methods, "Drug Administration"). All 195 concentrations are troughs drawn
about 30 min before the morning dose, first sampled a median 4 days after
transplantation (Table 1). The fit used NONMEM 7.4 with FOCE-I.

The same information is stored in the model's `population` metadata.

# Source trace

| Model element | Value | Source location |
|---|---|---|
| Structure | one-compartment, IV input, first-order elimination | Methods, "Population Pharmacokinetic Modeling" |
| `lcl` | log(2.08 L/h) | Table 2 final model, CL |
| `lvc` | log(63.2 L) | Table 2 final model, V |
| `e_crcl_cl` | 0.698 | Table 2, theta_1 (GFR on CL) |
| `e_wt_cl` | 1.07 | Table 2, theta_2 (WT on CL) |
| `e_wt_vc` | 0.934 | Table 2, theta_3 (WT on V) |
| CL equation | `2.08 * (WT/59.95)^1.07 * (GFR/36.67)^0.698` | Results, "Assessment of Covariates and Evaluation of Models" |
| V equation | `63.2 * (WT/59.95)^0.934` | Results, same section |
| `etalcl` | log(0.215^2 + 1) = 0.04517 | Table 2, omega_1 = 21.5% (CV reading; see below) |
| IIV on V | none | Table 2 prints no omega for V |
| `propSd` | 0.242 | Table 2, sigma_1 = 24.2%; Results: proportional model selected |
| Validation target | 15 weight/GFR daily-dose regimens, each >90% AUC0-24 >= 400 | Table 3; Methods, "Model Application" |

## The omega scale

Table 2 heads its IIV row "Intersubject variance of CL" and prints 21.5%. That
could be a coefficient of variation (omega^2 = log(0.215^2 + 1) = 0.0452) or
the variance itself (omega^2 = 0.215, a CV of about 49%). Table 3 separates the
two. Its 15 daily-dose recommendations are each reported to give more than 90%
of patients an AUC0-24 of at least 400 mg*h/L (MIC 1 mg/L). At steady state
AUC0-24 = daily dose / CL exactly, and CL is log-normal, so the fraction reaching
the target is available in closed form:
`P(AUC >= 400) = pnorm(log(dose / (400 * CL_typ)) / omega)`.

```{r table3}
# Ma 2020 Table 3, transcribed.
table3 <- expand.grid(GFR = c(30, 45, 60, 75, 90), WT = c(40, 50, 60)) |>
  dplyr::select(WT, GFR) |>
  dplyr::mutate(
    daily_dose = c(
      750, 1000, 1250, 1500, 1750,
      1000, 1250, 1500, 1750, 2000,
      1250, 1500, 1750, 2250, 2500
    ),
    scenario = sprintf("WT %d kg, GFR %d mL/min", WT, GFR)
  )

cl_typ <- function(wt, gfr) 2.08 * (wt / 59.95)^1.07 * (gfr / 36.67)^0.698

pta_closed <- table3 |>
  dplyr::mutate(
    cl_typical = cl_typ(WT, GFR),
    pta_cv = pnorm(log(daily_dose / (400 * cl_typical)) / sqrt(log(1 + 0.215^2))),
    pta_variance = pnorm(log(daily_dose / (400 * cl_typical)) / sqrt(0.215))
  )

pta_closed |>
  dplyr::select(WT, GFR, daily_dose, cl_typical, pta_cv, pta_variance) |>
  dplyr::mutate(dplyr::across(c(pta_cv, pta_variance), ~ 100 * .x)) |>
  dplyr::rename(
    "WT (kg)" = WT,
    "GFR (mL/min)" = GFR,
    "Daily dose (mg)" = daily_dose,
    "Typical CL (L/h)" = cl_typical,
    "% AUC0-24 >= 400, CV reading" = pta_cv,
    "% AUC0-24 >= 400, variance reading" = pta_variance
  ) |>
  knitr::kable(digits = c(0, 0, 0, 2, 1, 1))

# Deterministic (closed form, no simulation). The CV reading reproduces the
# paper's '>90%' in every cell; the variance reading fails every cell.
stopifnot(
  all(pta_closed$pta_cv > 0.90),
  all(pta_closed$pta_variance < 0.90)
)
```

The CV reading gives 97-99.5% in every cell and the variance reading gives only
80-88%, so the variance reading contradicts every row of Table 3. The model
uses the CV reading. The residual row (24.2%) is read on the same scale as a
proportional SD of 0.242.

# Virtual cohort

Each of the 15 Table 3 scenarios is simulated as its own arm of 200 subjects
with weight and GFR fixed at the scenario values. The daily dose is split into
two 12-hourly 1-hour infusions (the paper gives neither the split nor the
infusion time; neither affects AUC0-24 at steady state) for 20 days, so the
last day is at steady state for every subject (the typical half-life is
11-26 h across the scenarios, and a subject with low clearance in the lightest,
lowest-GFR scenario reaches about 45 h; 10 days would leave such subjects a few
percent short of steady state).

```{r cohort}
# rxSetSeed() fixes the draw within an rxode2 build; the assertions below are
# written to hold for any cohort the model can produce.
rxode2::rxSetSeed(2020)

mod <- readModelDb("Ma_2020_vancomycin")
n_per_arm <- 200L
tau <- 12
n_doses <- 40L
t_inf <- 1
t_last <- n_doses * tau

make_arm <- function(i) {
  row <- table3[i, ]
  ids <- (i - 1L) * n_per_arm + seq_len(n_per_arm)
  dose_rows <- expand.grid(id = ids, time = (seq_len(n_doses) - 1L) * tau) |>
    dplyr::mutate(evid = 1L, amt = row$daily_dose / 2, dur = t_inf)
  obs_times <- sort(unique(c(seq(0, t_last, by = 2), seq(t_last - 24, t_last, by = 0.25))))
  obs_rows <- expand.grid(id = ids, time = obs_times) |>
    dplyr::mutate(evid = 0L, amt = 0, dur = 0)
  dplyr::bind_rows(dose_rows, obs_rows) |>
    dplyr::mutate(
      cmt = "central",
      WT = row$WT,
      CRCL = row$GFR,
      scenario = row$scenario
    ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(lapply(seq_len(nrow(table3)), make_arm))
stopifnot(dplyr::n_distinct(events$id) == n_per_arm * nrow(table3))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod, events,
  keep = c("scenario", "WT", "CRCL"),
  returnType = "data.frame"
)
```

# Replicate published figures

The paper's figures are observed data and diagnostics (Figure 1, observed
troughs; Figures 3-5, goodness of fit and NPDE), so no model-predicted figure is
available to overlay. The figure below shows the simulated median and 90%
interval of the concentration-time profile over the final (steady-state) day for
the 60 kg scenarios, the heaviest weight band in Table 3.

```{r figure-profiles, fig.width = 7, fig.height = 4}
sim |>
  dplyr::filter(WT == 60, time >= t_last - 24) |>
  dplyr::group_by(scenario, time) |>
  dplyr::summarise(
    q05 = quantile(Cc, 0.05), q50 = median(Cc), q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time - (t_last - 24), q50, colour = scenario, fill = scenario)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.15, colour = NA) +
  geom_line() +
  labs(
    x = "Time within day 20 (h)", y = "Vancomycin (mg/L)",
    colour = NULL, fill = NULL,
    title = "Table 3 regimens, 60 kg: day-20 profile (median, 90% interval)"
  ) +
  theme_bw()
```

# Deterministic structural checks

With random effects zeroed, the model's clearance and volume must equal the
Results equations exactly.

```{r gate-equations}
mod_typical <- rxode2::zeroRe(mod)
ev_typ <- events |>
  dplyr::filter(id %in% ((seq_len(nrow(table3)) - 1L) * n_per_arm + 1L))
sim_typ <- rxode2::rxSolve(
  mod_typical, ev_typ,
  keep = c("scenario", "WT", "CRCL"),
  returnType = "data.frame"
)
chk_eq <- sim_typ |>
  dplyr::distinct(scenario, WT, CRCL, cl, vc) |>
  dplyr::mutate(
    cl_paper = cl_typ(WT, CRCL),
    vc_paper = 63.2 * (WT / 59.95)^0.934
  )
stopifnot(nrow(chk_eq) == nrow(table3))
stopifnot(
  max(abs(chk_eq$cl / chk_eq$cl_paper - 1)) < 1e-10,
  max(abs(chk_eq$vc / chk_eq$vc_paper - 1)) < 1e-10
)
```

# PKNCA validation

AUC is computed with PKNCA over the final day (456-480 h) for every simulated
subject, grouped by scenario.

```{r pknca}
conc_df <- sim |>
  dplyr::filter(!is.na(Cc), time >= t_last - 24) |>
  dplyr::select(id, time, Cc, scenario)

dose_df <- events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, scenario)

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | scenario + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | scenario + id)
intervals <- data.frame(
  start = t_last - 24, end = t_last,
  auclast = TRUE, cmax = TRUE, cmin = TRUE
)
nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)

nca_wide <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("auclast", "cmax", "cmin")) |>
  dplyr::select(scenario, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

# A gate that cannot go red is worse than none: every subject present, every
# AUC resolved.
stopifnot(nrow(nca_wide) == n_per_arm * nrow(table3), !anyNA(nca_wide$auclast))
```

## Steady-state AUC against Dose / CL

Each subject's day-20 AUC should equal its daily dose divided by its own
clearance. The two sides use the same drawn parameters, so the difference is
only residual accumulation and trapezoidal error, and a tight bound applies.

```{r gate-auc-dose-cl}
cl_ind <- sim |>
  dplyr::distinct(id, cl)
chk_auc <- nca_wide |>
  dplyr::left_join(cl_ind, by = "id") |>
  dplyr::left_join(dplyr::select(table3, scenario, daily_dose), by = "scenario") |>
  dplyr::mutate(ratio = auclast * cl / daily_dose)
stopifnot(!anyNA(chk_auc$ratio))
stopifnot(max(abs(chk_auc$ratio - 1)) < 0.01)
```

# Comparison against the published target attainment (Table 3)

The paper reports no NCA summary; its model-derived results are the Table 3
target-attainment percentages. The simulated fraction of each 200-subject arm
reaching AUC0-24 >= 400 mg*h/L is compared with the paper's ">90%". The typical
AUC0-24 (daily dose / typical CL, which is the median of a log-normal AUC) is
the reference for the simulated median AUC.

```{r nca-compare}
auc_ref <- pta_closed |>
  dplyr::transmute(scenario, auclast = daily_dose / cl_typical)

sim_summary <- nca_wide |>
  dplyr::select(scenario, id, auclast)

ncaComparisonTable(
  simulated = sim_summary,
  reference = auc_ref,
  by = "scenario",
  params = "auclast",
  units = c(auclast = "mg*h/L")
) |>
  knitr::kable()

pta_sim <- nca_wide |>
  dplyr::group_by(scenario) |>
  dplyr::summarise(pta = mean(auclast >= 400), .groups = "drop") |>
  dplyr::left_join(dplyr::select(pta_closed, scenario, pta_cv), by = "scenario")

pta_sim |>
  dplyr::mutate(pta = 100 * pta, pta_cv = 100 * pta_cv, paper = ">90%") |>
  dplyr::rename(
    "Scenario" = scenario,
    "Simulated % AUC0-24 >= 400" = pta,
    "Closed-form %" = pta_cv,
    "Ma 2020 Table 3" = paper
  ) |>
  knitr::kable(digits = 1)
```

```{r gate-table3}
stopifnot(nrow(pta_sim) == nrow(table3))
# Centre: the simulated medians track the closed-form typical AUC.
med_diff <- nca_wide |>
  dplyr::group_by(scenario) |>
  dplyr::summarise(med = median(auclast), .groups = "drop") |>
  dplyr::left_join(auc_ref, by = "scenario") |>
  dplyr::mutate(pct = 100 * (med / auclast - 1))
stopifnot(abs(median(med_diff$pct)) < 5)
# Table 3. Each cell's expected attainment is 97-99.5% (closed form above);
# with 200 subjects the binomial SE is about 1.2 percentage points, so the
# worst cell sits about 6 SE above the paper's 90% line. A mis-transcribed
# exponent, reference value or omega moves these by far more.
stopifnot(median(pta_sim$pta) > 0.95, all(pta_sim$pta > 0.90))
```

The simulated attainment reproduces the paper's ">90%" in every cell.

# Assumptions and deviations

* **Omega scale.** Table 2's "Intersubject variance of CL = 21.5%" is read as a
  CV and converted to omega^2 = log(0.215^2 + 1). The variance reading fails
  every row of Table 3 (see "The omega scale"). Table 2's residual "24.2%" is
  read as a proportional SD of 0.242 on the same basis; no source value
  discriminates the residual scale, but it does not affect AUC-based targets.
* **GFR equation.** The typeset GFR formula in the Methods has lost its exponent
  formatting ("GFR = 2.104*sCr(uM) - 1.154*age - 1.154*(0.742 for female)*1.233";
  cited to Chu et al. 2020), so the exact estimating equation cannot be recovered
  from this paper. The model takes the GFR estimate directly as the `CRCL`
  column in raw mL/min (not BSA-normalized).
* **GFR reference value.** The CL equation normalizes by 36.67 mL/min, which is
  not the Table 1 per-patient median (39.91 mL/min). The paper does not say
  what 36.67 is (possibly a median over observation records). The equation value
  is used.
* **Time-varying covariates.** Figures 2B and 2D show body weight and GFR
  changing over the postoperative period; the paper does not say whether they
  entered the model as time-varying. Both are accepted as time-varying columns.
* **Infusion duration and dose split.** Not reported. The vignette uses 1-hour
  infusions every 12 hours. Neither choice affects AUC0-24 at steady state, which
  is what Table 3 reports.
* **IIV on V.** Table 2 reports no IIV on V (the trough-only design gives little
  information on V), so none is included.
