---
title: "Difloxacin in crucian carp (Ma 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Difloxacin in crucian carp (Ma 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Ma_2023_difloxacin_crucianCarp"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Article: <https://doi.org/10.3390/vetsci10070416>\n")
cat("* Open access full text (PMC): <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10383650/>\n")
```

`r ui$description`

Ma 2023 is, in the authors' words, the first population PK analysis of
difloxacin in crucian carp. Difloxacin is a third-generation fluoroquinolone
used extra-label in Chinese aquaculture; no fish-specific dosing guidance
exists, so mammalian regimens are extrapolated. The paper's practical
conclusion is that the 20 mg/kg oral dose in current use does not reach a free
AUC/MIC of 125 against *Aeromonas hydrophila*.

# Population

Thirty healthy crucian carp (*Carassius auratus*) were split into six equal
groups of five; one group was an undosed control supplying blank plasma,
leaving **25 dosed fish** (Section 2.2). Mean body weight was 0.298 kg
(range 0.21 to 0.40 kg; per-fish weights are listed in Table 1). Fish were held
at 21.3 +/- 1.2 degC after at least 10 days of acclimation, in water at pH ~7.4
with dissolved oxygen above 7.5 mg/L.

Each fish received a single 20 mg/kg body-weight oral gavage dose of difloxacin
hydrochloride (10 mg/mL in 0.9% saline) and was **sampled sparsely, three times
only**, from the tail vein. Table 1 shows five repeating three-point schedules
cycling across the 25 fish, between them covering all 15 nominal sampling
times from 0.25 to 120 h. Plasma difloxacin was measured by HPLC-UV at 276 nm
(calibration 0.1 to 5 ug/mL, LOQ 0.1 ug/mL).

The same information is available programmatically via the model's `population`
metadata:

```{r population}
str(ui$population)
```

## Study design (Ma 2023 Table 1)

The sparse-sampling design is reproduced below exactly as printed, and is used
later to evaluate the model on the paper's own measurement grid rather than on
a finer grid of our choosing.

```{r table1}
# Ma 2023 Table 1: per-fish body weight and the three sampling times.
# The five schedules cycle in order across fish 1..25.
schedules <- list(
  A = c(0.25, 6, 36),
  B = c(0.5, 8, 48),
  C = c(1, 12, 72),
  D = c(2, 16, 96),
  E = c(4, 24, 120)
)
table1 <- tibble::tibble(
  id = 1:25,
  # Ma 2023 Table 1, BW (kg) column, in printed ID order
  WT = c(0.21, 0.33, 0.37, 0.35, 0.31, 0.26, 0.28, 0.26, 0.27, 0.27,
         0.30, 0.27, 0.34, 0.23, 0.40, 0.25, 0.28, 0.27, 0.29, 0.36,
         0.33, 0.29, 0.29, 0.34, 0.31),
  schedule = rep(names(schedules), times = 5)
)

# The 15 nominal times of Section 2.3, each covered by exactly 5 fish.
nominal_times <- c(0.25, 0.5, 1, 2, 4, 6, 8, 12, 16, 24, 36, 48, 72, 96, 120)
covered <- sort(unlist(schedules, use.names = FALSE))
stopifnot(
  nrow(table1) == 25L,
  # Section 2.2: 30 fish in 6 groups of 5, one group is the undosed control.
  nrow(table1) == 30L - 5L,
  # Section 2.3: "Each fish was sparsely sampled three times".
  all(lengths(schedules) == 3L),
  setequal(covered, nominal_times),
  # Weight range and mean quoted in Section 2.2.
  min(table1$WT) == 0.21, max(table1$WT) == 0.40,
  abs(mean(table1$WT) - 0.298) < 0.005
)
c(n_fish = nrow(table1), n_samples = nrow(table1) * 3L,
  mean_WT = round(mean(table1$WT), 3))
```

The printed Table 1 carries 74 asterisks rather than 75: fish 1 shows marks
only at 6 and 36 h. Section 2.3 states unambiguously that *each* fish was
sampled three times, and every other fish following schedule A carries the
0.25 h mark, so the missing mark is read here as a typesetting omission.

# Source trace

Every value below also carries an in-file comment next to its `ini()` entry in
`inst/modeldb/specificDrugs/Ma_2023_difloxacin_crucianCarp.R`.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lka` (tvKa) | 1.18 1/h | Table 3, **Final model (Covariance model)** row; RSE 12.55%, CI 0.88 to 1.47 |
| `lvc` (tvV, V/F) | 14.18 L/kg | Table 3, Final model row; RSE 18.88%, CI 8.83 to 19.53 |
| `lcl` (tvCL, CL/F) | 0.20 L/h/kg | Table 3, Final model row; RSE 17.96%, CI 0.13 to 0.27 |
| `etalka` variance | 0.69 | Table 3, "Inter-Individual Variation omega^2 (RSE%)" column, Final model; RSE 35.14% |
| `etalvc` variance | 1.12 | Table 3, same column, Final model; RSE 28.54% |
| `etalcl` variance | 0.98 | Table 3, same column, Final model; RSE 27.08% |
| `expSd` (stdev0) | 0.16 | Table 3, Final model; RSE 11.06%, CI 0.12 to 0.19. Log-additive error model per Section 2.5 |
| One compartment, first-order oral absorption, no lag | n/a | Section 2.5 (Ka / V / CL parameterisation, Phoenix NLME 8.1); Equations 1 to 3 |
| Body weight excluded as a covariate | n/a | Section 2.5 and Section 3.3: delta -2LL = 3.02 < 6.64 |
| Dose 20 mg/kg BW, single oral gavage | n/a | Section 2.3 |

Values in the packaged model come from the **final** (covariance) model, not the
basic model. For reference the basic model gave tvKa 1.16, tvV 14.34,
tvCL 0.18, stdev0 0.19.

## The residual error model

Section 2.5 states that "the log-additive error model was used". Phoenix NLME's
log-additive model is `CObs = C * exp(eps)` with `eps ~ N(0, stdev0^2)`, i.e. an
additive residual on the natural-log concentration. That is exactly nlmixr2's
`~ lnorm(expSd)`, so `expSd = 0.16` is a **dimensionless log-scale SD**. Table 3
prints "ug/mL" in the Units column for `stdev0`; that column also mis-assigns
units across the merged parameter rows, and a log-additive SD cannot carry
concentration units, so the printed unit is treated as a reporting artefact.

# Reproducing the paper's secondary parameters (Table 4)

Section 2.5 defines the secondary parameters by three closed forms:

* Equation 1: `K10 = tvCL / tvV`
* Equation 2: `T1/2 = ln(2) * tvV / tvCL`
* Equation 3: `AUC0-inf = Dose / tvCL`

Table 4 reports K10 = 0.01 1/h, T1/2 = 50.14 h and AUC0-inf = 102.05 h*ug/mL.
Solving the packaged ODE model at its typical values and running PKNCA on the
result is a genuine end-to-end check: it exercises the ODE structure, the
dosing route, the body-weight-normalised units and the transcribed parameter
values against three published numbers.

```{r typical-solve}
mod <- readModelDb("Ma_2023_difloxacin_crucianCarp")
mod_typical <- rxode2::zeroRe(mod)

# Fine early grid to resolve Tmax (pattern 11: a coarse grid understates AUC),
# then out to 480 h -- nearly 10 elimination half-lives -- so the aucinf
# extrapolation is a small correction rather than the bulk of the estimate.
grid_typical <- sort(unique(c(seq(0, 24, by = 0.05), seq(24, 480, by = 1))))
ev_typical <- rxode2::et(amt = 20, cmt = "depot") |>
  rxode2::et(grid_typical)

sim_typical <-
  rxode2::rxSolve(mod_typical, ev_typical, returnType = "data.frame") |>
  mutate(id = 1L, treatment = "20 mg/kg")

# Concentrations must stay non-negative or PKNCA's lambda.z log() misbehaves.
stopifnot(all(sim_typical$Cc >= 0))
```

```{r typical-profile, echo = FALSE}
sim_typical |>
  filter(time <= 120) |>
  ggplot(aes(time, Cc)) +
  geom_line(linewidth = 0.8) +
  scale_x_continuous(breaks = c(0, 12, 24, 48, 72, 96, 120)) +
  labs(
    x = "Time (h)", y = "Plasma difloxacin (ug/mL)",
    title = "Typical-value profile, single 20 mg/kg oral dose",
    caption = "Companion to Figure 1 of Ma 2023 (which plots the individual observed data)."
  ) +
  theme_bw()
```

# PKNCA validation

```{r pknca}
nca_conc <- sim_typical |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time = 0 record (extravascular pre-dose Cc = 0).
nca_conc <- dplyr::bind_rows(
  nca_conc,
  nca_conc |> dplyr::distinct(id, treatment) |>
    dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ time | treatment + id)

dose_df <- data.frame(id = 1L, time = 0, amt = 20, treatment = "20 mg/kg")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE,
  aucinf.obs = TRUE, half.life = TRUE, lambda.z = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
nca_wide[, c("cmax", "tmax", "auclast", "aucinf.obs", "half.life", "lambda.z")]
```

## Gate 1 -- Table 4 secondary parameters

```{r gate-table4}
auc_sim   <- nca_wide$aucinf.obs
thalf_sim <- nca_wide$half.life
k10_sim   <- nca_wide$lambda.z

pct <- function(sim, ref) (sim - ref) / ref * 100

# Both quantities are DETERMINISTIC (zeroRe: no cohort is drawn), so the
# tolerance is a transcription tolerance, not a sampling tolerance.
#
# Realised: -2.01% on AUC and -2.00% on T1/2, both traceable to a single
# cause -- Table 3 prints tvCL rounded to 2 dp as 0.20, while Table 4's
# AUC0-inf = 102.05 was computed from the unrounded 20/102.05 = 0.19598.
# The packaged model uses the printed 0.20 (see "Assumptions and
# deviations"). A 3% band admits that rounding and nothing else: a 10%
# error in a transcribed clearance, volume, dose or unit moves both
# numbers by 10% or more and turns this red.
stopifnot(
  abs(pct(auc_sim, 102.05)) < 3,
  abs(pct(thalf_sim, 50.14)) < 3,
  # Table 4 prints K10 to 2 dp; the model must round to the same value.
  round(k10_sim, 2) == 0.01
)

tibble::tibble(
  Quantity  = c("AUC0-inf (h*ug/mL)", "T1/2 (h)", "K10 (1/h)"),
  `Ma 2023 Table 4` = c(102.05, 50.14, 0.01),
  Simulated = round(c(auc_sim, thalf_sim, k10_sim), c(2, 2, 4)),
  `% diff`  = round(pct(c(auc_sim, thalf_sim, k10_sim), c(102.05, 50.14, 0.01)), 2)
) |>
  knitr::kable(caption = "Gate 1: PKNCA on the typical-value solve vs Ma 2023 Table 4.")
```

The residual 2% is fully explained by the printed rounding of `tvCL`. Solving
for the unrounded values that Table 4 implies confirms both round back to the
values Table 3 prints, so the two tables are internally consistent and no
transcription error is hiding in the gap:

```{r rounding-interval}
cl_unrounded <- 20 / 102.05                       # Equation 3 inverted
v_unrounded  <- 50.14 * cl_unrounded / log(2)     # Equation 2 inverted
stopifnot(
  round(cl_unrounded, 2) == 0.20,   # Table 3 prints 0.20
  round(v_unrounded, 2) == 14.18    # Table 3 prints 14.18
)
c(tvCL_implied = round(cl_unrounded, 5), tvV_implied = round(v_unrounded, 4))
```

## Gate 2 -- Tmax on the paper's own sampling grid

Table 2 reports Tmax = 4.00 h from the naive-averaged data. That value can only
ever be one of the 15 nominal sampling times, so the correct comparison
restricts the model to the paper's measurement grid rather than to the fine
grid used above.

```{r gate-tmax}
on_grid <- sim_typical |> dplyr::filter(time %in% nominal_times)
stopifnot(nrow(on_grid) == length(nominal_times))
tmax_grid <- on_grid$time[which.max(on_grid$Cc)]

# Exact: the model's peak must land on the same nominal time the paper
# reports. The true continuous Tmax is 3.80 h, and 4 h is its nearest
# neighbour on the grid; a mis-transcribed Ka moves the peak to 2 or 6 h.
stopifnot(tmax_grid == 4)
c(tmax_on_paper_grid = tmax_grid, tmax_continuous = round(nca_wide$tmax, 2))
```

## Comparison against the published NCA

```{r nca-compare}
# Ma 2023 reports two NCA-style parameter sets. Table 4 holds the secondary
# parameters DERIVED FROM THE FINAL MODEL, which is the like-for-like
# reference for a simulation from that model. Table 2's Cmax and Tmax are
# the only published values for those two parameters and come from a
# naive-averaged-data NCA, which the authors used only to seed initial
# estimates (Discussion, paragraph 1) -- they are shown here for context and
# are expected to differ.
published <- tibble::tribble(
  ~treatment,  ~cmax, ~tmax, ~aucinf.obs, ~half.life,
  "20 mg/kg",  1.13,  4.00,  102.05,      50.14
)

simulated <- nca_wide |>
  dplyr::mutate(treatment = "20 mg/kg") |>
  dplyr::select(treatment, cmax, tmax, aucinf.obs, half.life)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = simulated,
  reference     = published,
  by            = "treatment",
  units         = c(cmax = "ug/mL", tmax = "h",
                    aucinf.obs = "h*ug/mL", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated (typical value) vs published NCA.",
    "AUC0-inf and T1/2 are from Table 4 (model-derived);",
    "Cmax and Tmax are from Table 2 (naive-averaged-data NCA).",
    "* differs from reference by >20%."
  ),
  align = c("l", "l", "r", "r", "r")
)
attr(cmp, "footnote")
```

`Cmax` is the one row that can approach the flag: the model's typical-value
peak is 1.34 ug/mL against 1.13 ug/mL read off the naive average of sparse
observations. The two are not the same quantity. The naive average pools five
different fish at each nominal time, so between-fish variability -- which here
is very large (omega^2 of 1.12 on V and 0.98 on CL) -- flattens and lowers the
averaged peak relative to the typical-value curve. Table 2's own AUC0-inf
(60.14) sits 41% below the model-derived AUC0-inf (102.05) for the same reason,
which is precisely why the authors treated the naive-averaged NCA as a source
of initial estimates only.

## An arithmetic error in Table 2

Table 2 prints CL = 0.03 L/h/kg for the naive-averaged NCA. That value is
inconsistent with every other entry in the same table. Its three companions
over-determine the clearance:

```{r table2-typo}
# Ma 2023 Table 2 (naive-averaged-data NCA)
t2 <- list(auc_t = 39.67, auc_inf = 60.14, v = 27.63, cl = 0.03,
           tmax = 4.00, cmax = 1.13, thalf = 57.59, pct_extrap = 34.03)
dose <- 20

cl_from_auc <- dose / t2$auc_inf                       # CL = Dose / AUC0-inf
lz          <- log(2) / t2$thalf                       # lambda.z from T1/2
v_recon     <- dose / (t2$auc_inf * lz)                # Vz = Dose / (AUC * lz)
extrap_recon <- (t2$auc_inf - t2$auc_t) / t2$auc_inf * 100

# Vz and %AUC extrapolated both reconstruct to the printed values from
# AUC0-inf and T1/2 alone, so those three entries are mutually consistent...
stopifnot(
  abs(v_recon - t2$v) / t2$v * 100 < 0.5,
  abs(extrap_recon - t2$pct_extrap) < 0.1
)
# ...and they imply CL = 0.33, not the printed 0.03: an order of magnitude out.
stopifnot(abs(cl_from_auc / t2$cl - 11) < 1)

c(CL_printed = t2$cl, CL_implied = round(cl_from_auc, 4),
  V_printed = t2$v, V_reconstructed = round(v_recon, 2),
  extrap_printed = t2$pct_extrap, extrap_reconstructed = round(extrap_recon, 2))
```

`Vz` reconstructs to 27.63 L/kg and the extrapolated fraction to 34.03%, both
exactly as printed, from `AUC0-inf` and `T1/2` alone. The same two numbers give
`CL = Dose / AUC0-inf = 0.333 L/h/kg`. The printed 0.03 is therefore a
digit-transposition typo for 0.33. This affects only Table 2's descriptive NCA;
**no packaged model parameter derives from it**, since the model uses Table 3.

# Between-fish variability and the visual predictive check

Ma 2023 Figure 4 is a VPC of the final model over 0 to 120 h with 1000
simulation replicates. The observed data are not published, so only the
predicted quantiles can be reproduced. The paper describes the between-fish
variability as "huge", and the packaged omegas bear that out.

```{r cohort}
# rxSetSeed() fixes rxode2's stream per solver thread but NOT across thread
# counts, so a CI runner draws a different cohort than a workstation. Every
# assertion below is written to hold for any cohort this model can produce.
rxode2::rxSetSeed(20230627)

n_fish <- 200L  # cap is 200 per arm
grid_cohort <- sort(unique(c(
  seq(0, 24, by = 0.25), seq(25, 120, by = 1), seq(124, 480, by = 4)
)))
ev_cohort <- rxode2::et(amt = 20, cmt = "depot") |>
  rxode2::et(grid_cohort) |>
  rxode2::et(id = seq_len(n_fish))

sim_cohort <-
  rxode2::rxSolve(mod, ev_cohort, returnType = "data.frame") |>
  mutate(treatment = "20 mg/kg")

stopifnot(dplyr::n_distinct(sim_cohort$id) == n_fish, all(sim_cohort$Cc >= 0))
```

```{r figure-4, echo = FALSE}
# Replicates Figure 4 of Ma 2023: 5th / 50th / 95th percentiles of the
# simulated dependent variable over the observation window. The `sim` column
# carries the log-normal residual error, matching the paper's DV quantiles;
# `Cc` is the individual prediction without residual error.
sim_cohort |>
  filter(time > 0, time <= 120) |>
  group_by(time) |>
  summarise(
    Q05 = quantile(sim, 0.05), Q50 = quantile(sim, 0.50),
    Q95 = quantile(sim, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(linewidth = 0.8) +
  scale_x_continuous(breaks = c(0, 12, 24, 48, 72, 96, 120)) +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Plasma difloxacin (ug/mL, log scale)",
    title = "Figure 4 -- simulated 5th / 50th / 95th percentiles",
    caption = paste(
      "Replicates the predicted-quantile bands of Figure 4 of Ma 2023;",
      n_fish, "simulated fish, single 20 mg/kg oral dose."
    )
  ) +
  theme_bw()
```

## Gate 3 -- the omega block reproduces as encoded

Per-fish `ka`, `vc` and `cl` are returned directly by `rxSolve`, so the
realised cohort can be checked against the encoded log-normal parameters
without any NCA step.

```{r gate-omega}
per_fish <- sim_cohort |>
  group_by(id) |>
  summarise(ka = first(ka), vc = first(vc), cl = first(cl), .groups = "drop")

realised <- tibble::tibble(
  parameter = c("ka", "vc", "cl"),
  encoded_median = c(1.18, 14.18, 0.20),
  encoded_omega  = sqrt(c(0.69, 1.12, 0.98)),
  cohort_median  = c(median(per_fish$ka), median(per_fish$vc), median(per_fish$cl)),
  cohort_sd_log  = c(sd(log(per_fish$ka)), sd(log(per_fish$vc)), sd(log(per_fish$cl)))
) |>
  mutate(
    median_log_ratio = log(cohort_median / encoded_median),
    sd_diff          = cohort_sd_log - encoded_omega
  )

# These are COHORT statistics, so the bounds must survive any draw.
# Measured over the three parameters at 1 / 2 / 4 / 8 / 16 solver threads:
#   max |log median ratio| = 0.156 / 0.028 / 0.126 / 0.068 / 0.247
#   max |SD difference|    = 0.033 / 0.063 / 0.058 / 0.084 / 0.073
# Theory agrees: SE(log median) ~= 1.2533 * omega / sqrt(n) = 0.094 at
# worst (vc), and SE(SD) ~= omega / sqrt(2n) = 0.053 at worst. The bounds
# below sit at ~6 SE, outside the whole observed range, and still go red
# on a decimal-point or variance-vs-SD error, which shift these by 1 or
# more. Do not tighten them back onto a single observed run.
stopifnot(
  max(abs(realised$median_log_ratio)) < 0.60,
  max(abs(realised$sd_diff)) < 0.30
)

realised |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  dplyr::rename(
    "Parameter" = parameter,
    "Encoded median" = encoded_median,
    "Encoded omega" = encoded_omega,
    "Cohort median" = cohort_median,
    "Cohort SD(log)" = cohort_sd_log,
    "log(median ratio)" = median_log_ratio,
    "SD difference" = sd_diff
  ) |>
  knitr::kable(caption = "Gate 3: realised cohort vs the encoded log-normal IIV.")
```

```{r gate-spread}
# The paper calls the between-fish variability "huge" (Discussion) and
# reports omega^2 above 0.9 for both V and CL. Gate 3 above checks the
# parameter columns; this checks that the variability actually propagates
# through the ODE solve and the residual-error draw into the simulated
# observations.
#
# The statistic is the INTERQUARTILE ratio, not a 5th/95th ratio. With
# omega^2 of 1.12 on V the far tail is extremely heavy and a P95/P05 ratio
# from 200 draws is useless as a gate: measured 61.3 / 15.0 / 19.0 / 18.6 /
# 89.8 at 1 / 2 / 4 / 8 / 16 solver threads, a six-fold swing on identical
# source. The IQR ratio over the same five runs was 3.08 / 3.21 / 3.19 /
# 2.80 / 3.55.
spread_24h <- sim_cohort |>
  filter(abs(time - 24) < 1e-8) |>
  summarise(ratio = quantile(sim, 0.75) / quantile(sim, 0.25)) |>
  pull(ratio)

# A bound of 2 sits below the whole observed 2.80-3.55 range and still goes
# red on the failure this gate exists to catch: with the etas dropped the
# only remaining spread is the residual error, giving an IQR ratio of
# exp(2 * 0.6745 * 0.16) = 1.24.
stopifnot(spread_24h > 2)
c(`P75:P25 ratio at 24 h` = round(spread_24h, 2))
```

# Gate 4 -- the free AUC/MIC conclusion (Discussion)

The paper's clinical conclusion is a short arithmetic chain built on the
model-derived AUC0-inf, a literature plasma protein-binding estimate and
literature MIC data. Running it from the **model's own** AUC rather than the
published 102.05 makes it an end-to-end check of the packaged model against a
published conclusion.

```{r gate-aucmic}
# Discussion, final paragraph:
#  - plasma protein binding of difloxacin in healthy gibel carp is 52.66% to
#    80.56%; the authors assume the mean, 66.7%, for crucian carp
#  - MIC of difloxacin against Aeromonas hydrophila isolated from carps:
#    0.83 to 4 ug/mL
#  - fluoroquinolone efficacy target: free AUC/MIC >= 125
fu <- 1 - 0.667
mic_range <- c(0.83, 4)
target_fauc_mic <- 125

free_auc_model <- auc_sim * fu
ratios_model   <- free_auc_model / mic_range

published_chain <- c(free_auc = 33.98, ratio_mic4 = 8.495, ratio_mic083 = 40.93)
model_chain     <- c(free_auc = free_auc_model,
                     ratio_mic4 = ratios_model[2], ratio_mic083 = ratios_model[1])

# Deterministic (typical value); the only slack is the 2% tvCL rounding
# already characterised in Gate 1, which propagates linearly through the
# whole chain.
stopifnot(all(abs((model_chain - published_chain) / published_chain * 100) < 3))

# The qualitative conclusion must survive: neither end of the MIC range
# reaches the efficacy target at 20 mg/kg.
stopifnot(all(ratios_model < target_fauc_mic))

tibble::tibble(
  Quantity = c("Free AUC0-inf (h*ug/mL)",
               "Free AUC/MIC at MIC 4 ug/mL",
               "Free AUC/MIC at MIC 0.83 ug/mL"),
  `Ma 2023 Discussion` = published_chain,
  Model = round(model_chain, 3),
  `% diff` = round((model_chain - published_chain) / published_chain * 100, 2)
) |>
  knitr::kable(caption = "Gate 4: the paper's free AUC/MIC chain, recomputed from the packaged model.")
```

Both ends of the reported MIC range fall well short of the free AUC/MIC target
of 125, reproducing the paper's conclusion that a single 20 mg/kg oral dose is
insufficient against *Aeromonas hydrophila* in crucian carp. Inverting the
target gives the highest MIC the dose could in principle cover:

```{r mic-breakpoint}
mic_covered <- free_auc_model / target_fauc_mic
c(`max MIC covered at free AUC/MIC 125 (ug/mL)` = round(mic_covered, 3))
```

0.27 ug/mL is below the lowest MIC (0.83 ug/mL) the authors cite, which is the
substance of the abstract's statement that the dose "cannot generate adequate
plasma concentrations to inhibit pathogens with MIC values above 0.83 ug/mL".

# Assumptions and deviations

* **The omega off-diagonals are not published.** Section 2.5 states that the
  final model used a *full variance-covariance matrix* over `etaV`, `etaCL` and
  `etaKa` -- indeed that is the only structural difference between the basic and
  final models, and it is what drove the reported delta -2LL of 42.72. Table 3
  publishes only the three diagonal variances; no covariance, correlation or
  full matrix appears anywhere in the paper or its figures. Rather than invent
  off-diagonal values, the packaged model leaves the three etas **independent**.
  Typical-value predictions (Gates 1, 2 and 4) are unaffected, because omega
  does not enter them. Cohort-level quantiles (the Figure 4 replication and
  Gate 3's spread check) will be somewhat narrower or wider than the paper's
  depending on the sign of the true correlations, most importantly the
  `etaCL`/`etaV` pair, which jointly determine the terminal slope.
* **`tvCL` is encoded as the printed 0.20 L/h/kg.** Table 4's secondary
  parameters were computed from an unrounded 0.19598 L/h/kg (recovered above by
  inverting Equations 2 and 3). Using the printed 2-dp value costs a uniform
  2.0% on AUC0-inf, T1/2 and every quantity derived from them. The printed value
  is used because it is what Table 3, the abstract, the results and the
  conclusions all state; the arithmetic to recover the unrounded value is shown
  in the "rounding-interval" chunk for anyone who prefers it.
* **`stdev0` is treated as a log-scale SD, not a concentration.** Justified in
  the "residual error model" section above from Section 2.5's explicit
  "log-additive error model".
* **Table 2's CL is a typo.** Demonstrated numerically above. No model parameter
  depends on Table 2.
* **Body weight is recorded but unused.** It is registered under
  `covariatesDataExcluded`, not `covariateData`, because Ma 2023 screened it and
  rejected it (delta -2LL = 3.02 against a 6.64 threshold). Note that weight
  scaling is nevertheless implicit in the units: the dose is mg/kg and V/F and
  CL/F are per kg, so the rejected test was for a *departure* from strict
  proportionality, not for the absence of any weight effect. Simulations must
  therefore supply `amt` in mg/kg body weight, and `Cc` comes back in ug/mL
  without any further scaling.
* **Observed data are not reproduced.** Ma 2023 publishes no concentration
  table, so Figures 1 to 4 cannot be overlaid with observations; only the
  model-side quantities are reproduced here. Table 1's design and body weights
  are transcribed exactly and are the only individual-level data in the paper.
* **Single temperature.** All parameters apply at 21.3 +/- 1.2 degC. The
  Discussion documents a roughly two-fold change in difloxacin half-life in
  crucian carp between 10 and 20 degC, so these values must not be extrapolated
  to other rearing temperatures.
* **No non-paper-derived parameter values.** Every `ini()` entry is transcribed
  from Ma 2023 Table 3; nothing was digitised from a figure, obtained by
  correspondence or carried from an upstream model.
