---
title: "IL-6-mediated CYP3A4 suppression (Machavaram 2013)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IL-6-mediated CYP3A4 suppression (Machavaram 2013)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 96
)
```

```{r libraries, message = FALSE, warning = FALSE}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(ggplot2)

mod <- rxode2::rxode2(readModelDb("Machavaram_2013_interleukin6_cyp3a_pbpk"))
```

## Scope of this model

Machavaram 2013 is the paper that introduced the cytokine enzyme-turnover
equation that the later IL-6 / CYP models in this library cite. Both
`Chen_2024_interleukin6_cyp3a_pbpk` and `Willemin_2024_interleukin6_cyp_pbpk`
attribute their turnover equation to it, and `Willemin_2024` additionally
carries this paper's IL-6 disposition parameters and its CYP3A4 potencies
unchanged. This vignette validates the founding form.

**What the model contains.** Two layers, both fully specified in the paper:

1. IL-6 disposition -- a one-compartment intravenous model, `CLiv` 1.0 L/h and
   `Vss` 0.43 L/kg, both estimated by fitting.
2. Hepatic CYP3A4 turnover -- the paper's Eq. 1, with the in vitro suppression
   constants `EC50` 73.2 pg/mL and `Emin` 0.24 and the Simcyp library
   degradation rate constant `kdeg` 0.0193/h.

**What the model deliberately does not contain.** The downstream victim-drug
predictions -- simvastatin exposure in rheumatoid arthritis (case study 1) and
cyclosporine exposure in bone marrow transplant patients (case studies 2 and 3)
-- are *not* reproduced here. Supplementary Tables S1 and S2 give the compound
layers (simvastatin `Vss` 2.26 L/kg, CYP3A4 `CLint` 2597 and additional
`CLint` 289 uL/min/mg microsomal protein; cyclosporine `Vss` 1.52 L/kg, CYP3A4
`CLint` 2.03 uL/min/pmol isoform), but turning either `CLint` into a systemic
clearance needs the microsomal protein per gram of liver, the liver weight, the
CYP3A4 abundance and the hepatic blood flow, and none of those is printed
anywhere on disk -- they are Simcyp V11.0 population-library outputs.

That gap is not cosmetic, and one number shows why. Cyclosporine is modelled
with all of its clearance through CYP3A4, so a naive well-stirred reduction
would predict an AUC ratio of `1 / 0.337` = 2.97 at a 500 pg/mL IL-6 clamp,
whereas the paper predicts a 45% AUC increase (ratio 1.45). The difference is
hepatic-blood-flow limitation, which cannot be reconstructed without `QH`.
Substituting a textbook `QH` would make the victim layer unauditable against
the source, so it is omitted rather than guessed.

The enzyme layer that *is* reproduced here is exactly what case study 4
validates against clinical data, and it is the layer every downstream
prediction in the paper is driven by.

## Population

The paper runs four Simcyp case studies plus a literature meta-analysis. No
subject-level data are fitted; these are virtual populations matched to the
demographics of the source clinical trials.

```{r population}
pop <- readModelDb("Machavaram_2013_interleukin6_cyp3a_pbpk")()$population

tibble::tibble(
  `Case study` = c("1", "2", "3", "4"),
  Population = c(
    "Rheumatoid arthritis", "Bone marrow transplant",
    "Bone marrow transplant", "Postsurgical trauma"
  ),
  Design = c(
    "10 trials x 12 subjects", "10 trials x 5 subjects",
    "10 trials x 1 subject", "10 trials x 16 subjects"
  ),
  Age = c("28-72 y", "18-38 y", "20 y and 34 y", "48-83 y"),
  `Female %` = c("66", "0", "0", "31"),
  `Victim drug` = c(
    "Simvastatin 40 mg oral", "Cyclosporine 1.5 mg/kg over 4 h",
    "Cyclosporine 1-3 mg/kg/day continuous", "None (enzyme endpoint only)"
  )
) |>
  knitr::kable(caption = "Case-study designs (Methods and Supplementary Methods).")
```

The meta-analysis (Results, Figure 1a) gives weighted geometric mean systemic
IL-6 of 4 pg/mL (95% CI 2-8) in healthy subjects, 54 pg/mL (43-69) in
rheumatoid arthritis and 229 pg/mL (174-300) postsurgery. Figures 1b and 1c
report the corresponding lognormal population distributions, and those are
reproduced as a gate below.

## Source trace

Every value in `ini()` and every equation in `model()`, with the place in the
source it came from.

```{r source-trace}
tibble::tribble(
  ~Quantity, ~Value, ~`Source location`,
  "IL-6 CL", "1.0 L/h",
  "Methods, 'Modeling of IL-6 profiles'; Supplementary Methods ('CLiv = 1.0 L/h')",
  "IL-6 Vss", "0.43 L/kg",
  "Methods, 'Modeling of IL-6 profiles'; Supplementary Methods",
  "IL-6 infusion-rate range", "0.0009-1.5 ug/h",
  "Methods, 'Modeling of IL-6 profiles'",
  "IL-6 target clamps", "1, 5, 10, 50, 100, 500, 1000, 1500 pg/mL",
  "Methods, 'Modeling of IL-6 profiles'; case study 4 uses 50/100/200/500",
  "IL-6 molecular weight", "21,000 g/mol",
  "Methods, 'Modeling of IL-6 profiles'",
  "Emin", "0.24",
  "Methods, 'Modeling of enzyme dynamics'; Supplementary Methods",
  "EC50", "73.2 pg/mL",
  "Methods, 'Modeling of enzyme dynamics'; Supplementary Methods",
  "kdeg (liver)", "0.0193 /h",
  "Methods, 'Modeling of enzyme dynamics', cited to Rowland Yeo 2011 (ref 38)",
  "Enzyme turnover equation", "Eq. 1",
  "Methods, Eq. 1; restated as Supplementary Methods Eq. 3",
  "IL-6 perturbation input", "Eq. 1 and Eq. 2",
  "Supplementary Methods, 'Modeling of IL-6 profiles'",
  "RA IL-6 distribution", "mu 3.99, sigma 1.2 (lognormal)",
  "Results, 'Meta-analysis and distribution of IL-6'; Figure 1b",
  "Postsurgery IL-6 distribution", "mu 5.4, sigma 1.0 (lognormal)",
  "Results, 'Meta-analysis and distribution of IL-6'; Figure 1c",
  "Enzyme baseline E0", "carried as 1 (relative)",
  "Eq. 1 definition (Enzact(t) = E0 at t = 0); no absolute abundance is printed",
  "IIV / residual error", "none reported",
  "Simulation-only paper; no variance, %CV or CI is given for any parameter"
) |>
  knitr::kable()
```

The turnover equation as printed in the paper is

$$
\frac{dEnz_{act,H-3A}}{dt} =
  k_{deg,H\text{-}3A} \cdot Enz_{0,H\text{-}3A}
  \left[ 1 + \frac{(E_{min} - 1)\,[I]_t}{EC_{50} + [I]_t} \right]
  - k_{deg,H\text{-}3A} \cdot Enz_{act,H\text{-}3A}
$$

which is transcribed verbatim into `model()`. IL-6 acts on the rate of enzyme
*synthesis* only; degradation is assumed unaffected by the perpetrator.

### Dimensional analysis

Amounts are mg and volumes L, so `central / vc` is mg/L. The paper reports
every IL-6 concentration, and `EC50`, in pg/mL, and 1 mg/L is exactly
1e6 pg/mL -- the `mgPerLToPgPerML` constant in `model()`. `Emin` is a
unitless fraction of vehicle control, `kdeg` is 1/h, and the enzyme state is a
unitless fraction of the untreated baseline, so `kdeg * fsupp` and
`kdeg * enzyme_3a4` are both 1/h and the equation is dimensionally consistent.

An independent check on the concentration scaling: at steady state under a
maintained infusion, `Css = R / CL`. The paper's largest rate is 1.5 ug/h and
its largest target clamp is 1500 pg/mL, so 1.5e-3 mg/h divided by 1.0 L/h is
1.5e-3 mg/L, which is 1500 pg/mL exactly. The reported rate range and the
reported concentration range therefore pin each other, and the check below
confirms the implementation reproduces it.

## Model structure

```{r structure}
mod
```

```{r helpers}
# Typical 70 kg subject; the model carries no IIV, so this is the whole model.
WT_REF <- 70
CL_REF <- 1.0                     # L/h, from ini()
V_REF <- 0.43 * WT_REF            # L, Vss 0.43 L/kg
KDEG <- 0.0193                    # 1/h
EMIN <- 0.24
EC50 <- 73.2

# Analytic steady-state enzyme activity, as a fraction of baseline. This is
# Eq. 1 with the derivative set to zero, so it is an independent closed form.
enz_ss <- function(il6) 1 + (EMIN - 1) * il6 / (EC50 + il6)

# Infusion rate (mg/h) that clamps systemic IL-6 to `css` pg/mL, from Css = R/CL.
rate_for_css <- function(css) css * 1e-6 * CL_REF

# Solve a clamped-IL-6 arm: a zero-order infusion held for `hours`.
solve_clamp <- function(css, hours, by = 6) {
  rate <- rate_for_css(css)
  ev <- rxode2::et(amt = rate * hours, rate = rate, cmt = "central") |>
    rxode2::et(seq(0, hours, by = by), cmt = "central")
  rxode2::rxSolve(mod, ev, params = c(WT = WT_REF), returnType = "data.frame")
}
```

## Check 1: with no IL-6 the enzyme pool holds at baseline

The paper states that in the absence of IL-6 the enzyme content is at steady
state, with synthesis equal to `kdeg * E0`. With `E0` carried as 1 the state
must sit at exactly 1 forever, and `fsupp` must be exactly 1.

```{r check-no-il6}
no_il6 <- rxode2::rxSolve(
  mod,
  rxode2::et(seq(0, 18 * 24, by = 6), cmt = "central"),
  params = c(WT = WT_REF), returnType = "data.frame"
)

max_drift <- max(abs(no_il6$enzyme_3a4 - 1))
max_conc <- max(no_il6$Cc)
c(max_enzyme_drift = max_drift, max_il6 = max_conc)

stopifnot(
  max_drift < 1e-8,
  max_conc < 1e-12
)
```

The pool is flat to within 1e-8 over 18 days. Both sides of this comparison are
deterministic, so a tight bound is the right bound.

## Check 2: the reported infusion rates reproduce the reported clamps

The paper reports an infusion-rate range (0.0009-1.5 ug/h) and a target
concentration range (1-1500 pg/mL) but never tabulates the pairing. `Css = R/CL`
implies it, and the top of both ranges must coincide exactly.

```{r check-css}
clamps <- c(1, 5, 10, 50, 100, 200, 500, 1000, 1500)

css_chk <- lapply(clamps, function(css) {
  s <- solve_clamp(css, hours = 18 * 24)
  tibble::tibble(
    target_pg_mL = css,
    rate_ug_h = rate_for_css(css) * 1000,
    achieved_pg_mL = dplyr::last(s$Cc)
  )
}) |>
  dplyr::bind_rows() |>
  dplyr::mutate(pct_diff = 100 * (achieved_pg_mL / target_pg_mL - 1))

knitr::kable(css_chk, digits = 4, caption = "Achieved IL-6 clamp vs target.")

stopifnot(
  # The implementation reproduces Css = R/CL to numerical precision.
  max(abs(css_chk$pct_diff)) < 0.01,
  # The paper's stated maximum rate of 1.5 ug/h must be the 1500 pg/mL clamp.
  abs(css_chk$rate_ug_h[css_chk$target_pg_mL == 1500] - 1.5) < 1e-9
)
```

The rate needed for the paper's top clamp is 1.5 ug/h, exactly the top of the
paper's reported rate range. The bottom of the reported range, 0.0009 ug/h,
corresponds to 0.9 pg/mL rather than the 1 pg/mL clamp -- a rounding of the
reported rate to one significant figure, not a model discrepancy.

## Check 3: steady-state suppression replicates case study 4 (Figure 6)

Case study 4 clamps systemic IL-6 at 50, 100, 200 and 500 pg/mL and compares
predicted hepatic CYP3A4 levels against erythromycin-breath-test activity in
postsurgical patients (Figure 6). The ODE solve must land on the closed form.

```{r check-enzyme-ss}
enz_chk <- lapply(clamps, function(css) {
  s <- solve_clamp(css, hours = 18 * 24)
  tibble::tibble(
    il6_pg_mL = css,
    solved_pct = 100 * dplyr::last(s$enzyme_3a4),
    closed_form_pct = 100 * enz_ss(css)
  )
}) |>
  dplyr::bind_rows() |>
  dplyr::mutate(pct_diff = 100 * (solved_pct / closed_form_pct - 1))

knitr::kable(
  enz_chk, digits = 3,
  caption = "Hepatic CYP3A4 activity (% of untreated baseline) at steady state."
)

stopifnot(
  # Pure numerical error between an ODE solve and its own closed form.
  max(abs(enz_chk$pct_diff)) < 0.1,
  # Emin is a floor: activity can never fall below 24% of baseline.
  all(enz_chk$solved_pct > 100 * EMIN),
  # Monotone decreasing in IL-6, which is the trend Figure 6 reports.
  all(diff(enz_chk$solved_pct) < 0)
)
```

The four case-study-4 clamps give 69.2%, 56.1%, 44.4% and 33.7% of baseline
hepatic CYP3A4. Figure 6 is a scatter plot without a printed value table, so
this reproduces its reported inverse trend and its "good agreement" claim
rather than a set of digits.

A useful cross-check on `EC50`: the paper defines it as the concentration
supporting half of the maximum suppressive effect. Maximum suppression is
`1 - Emin` = 0.76, half of that is 0.38, so activity at `EC50` must be 62%.

```{r check-ec50-definition}
at_ec50 <- 100 * enz_ss(EC50)
at_ec50
stopifnot(abs(at_ec50 - 62) < 0.01)
```

The paper's own three constants are internally consistent with its own
definition of `EC50`.

The very low clamps matter too. Case study 1 reports "minimal change" in
simvastatin pharmacokinetics at the 1-10 pg/mL levels seen in healthy
volunteers, and the enzyme layer must be correspondingly quiet there.

```{r check-low-clamps}
low <- dplyr::filter(enz_chk, il6_pg_mL <= 10)
knitr::kable(low, digits = 3, caption = "Healthy-volunteer IL-6 range.")
stopifnot(all(low$solved_pct > 90))
```

Suppression is under 10% across the healthy range, consistent with the paper's
conclusion that low IL-6 has negligible impact on CYP3A activity.

## Check 4: turnover half-life and time to the new steady state

The enzyme pool relaxes with a time constant of `1 / kdeg`, so the *terminal*
rate constant of its approach must be `kdeg` and the corresponding half-life
`log(2) / kdeg` = 35.9 h, regardless of the IL-6 level. The paper relies on this
when it states that a new hepatic CYP3A4 steady state is reached over a 10-day
simulation.

Two exponentials are in play and the check has to respect that. Starting an
infusion does not step IL-6 to its clamp: IL-6 accumulates with its own
half-life of `log(2) * V / CL` = 20.9 h at 70 kg, so the enzyme's approach is a
convolution of that ramp with its own turnover, and the time to cover half the
distance to the new steady state is 56 h, not 35.9 h. Because `kdeg`
(35.9 h) is the *slower* of the two processes, it governs the terminal phase, so
the right test is a log-linear regression on the residual once the IL-6 ramp has
died away.

```{r check-halflife}
s100 <- solve_clamp(100, hours = 30 * 24, by = 0.5)
target <- enz_ss(100)

# Terminal rate constant of the approach to the new steady state, measured
# after the 20.9 h IL-6 accumulation transient is spent.
window <- dplyr::filter(s100, time >= 150, time <= 500)
slope <- stats::coef(stats::lm(log(enzyme_3a4 - target) ~ time, data = window))
k_obs <- -unname(slope[2])

# Fraction of the way to the new steady state at 10 days.
at_10d <- dplyr::last(s100$enzyme_3a4[s100$time <= 240])
frac_10d <- (1 - at_10d) / (1 - target)

c(
  k_terminal_observed = k_obs,
  kdeg_declared = KDEG,
  t_half_observed_h = log(2) / k_obs,
  t_half_theory_h = log(2) / KDEG,
  fraction_complete_at_10d = frac_10d
)

stopifnot(
  # The terminal rate constant must recover the declared kdeg.
  abs(k_obs / KDEG - 1) < 0.02,
  # "a new steady-state level of CYP3A4 in the liver was achieved" over 10 days.
  frac_10d > 0.98
)
```

The recovered terminal rate constant is 0.01921/h against a declared `kdeg` of
0.0193/h, a half-life of 36.1 h against 35.9 h, and the pool is 98.3% of the way
to its new steady state at day 10. That is what makes the paper's simulation
design sound: the victim drug is dosed on day 10 of an 11-day run (case study 2)
or day 15 of an 18-day run (case study 1), by which point the enzyme layer has
settled.

```{r fig-timecourse, fig.cap = "Hepatic CYP3A4 activity after starting IL-6 clamps of 50-500 pg/mL, with the analytic steady state as a dashed line."}
tc <- lapply(c(50, 100, 200, 500), function(css) {
  s <- solve_clamp(css, hours = 20 * 24, by = 1)
  tibble::tibble(
    time = s$time, activity = 100 * s$enzyme_3a4,
    il6 = factor(css, levels = c(50, 100, 200, 500))
  )
}) |>
  dplyr::bind_rows()

ss_lines <- tibble::tibble(
  il6 = factor(c(50, 100, 200, 500), levels = c(50, 100, 200, 500)),
  activity = 100 * enz_ss(c(50, 100, 200, 500))
)

ggplot(tc, aes(time / 24, activity, colour = il6)) +
  geom_line(linewidth = 0.7) +
  geom_hline(
    data = ss_lines, aes(yintercept = activity, colour = il6),
    linetype = "dashed", linewidth = 0.4
  ) +
  labs(
    x = "Time (days)", y = "Hepatic CYP3A4 activity (% of baseline)",
    colour = "IL-6 (pg/mL)"
  ) +
  theme_bw()
```

## Check 5: the case-study-3 perturbation input

For case study 3 the paper did not clamp IL-6. It fitted each patient's
observed profile with a step from baseline synthesis `R0` (assumed zero) to a
perturbed rate `R1` held for a duration `T`, after which levels decline
exponentially. Supplementary Methods Eq. 1 and Eq. 2 give the closed form:

$$
C(t) = \frac{R_1}{CL}\left(1 - e^{-\frac{CL}{V}t}\right), \quad t \le T
\qquad
C(t) = \frac{R_1}{CL}\left(1 - e^{-\frac{CL}{V}T}\right)e^{-\frac{CL}{V}(t-T)},
\quad t > T
$$

That is exactly a zero-order infusion of rate `R1` and duration `T` into a
one-compartment model, so it needs no extra model structure -- it is an event
table. The gate below confirms the equivalence.

`R1` and `T` are *not* published for either patient, so they cannot be
reproduced. What *is* published is the observed IL-6 peak range for patients 1
and 5, 440-1228 pg/mL. The two arms below back-solve `R1` to hit each end of
that published range at an assumed `T` of 7 days; `R1` and `T` are therefore
illustrative, and only the peaks they reproduce are source values.

```{r check-perturbation}
T_DUR <- 7 * 24                              # assumed, not published
peaks <- c(440, 1228)                        # published observed peak range

# R1 that makes the analytic peak equal `peak`.
r1_for_peak <- function(peak) {
  peak * 1e-6 * CL_REF / (1 - exp(-CL_REF / V_REF * T_DUR))
}

pert <- lapply(peaks, function(peak) {
  r1 <- r1_for_peak(peak)
  ev <- rxode2::et(amt = r1 * T_DUR, rate = r1, cmt = "central") |>
    rxode2::et(seq(0, 19 * 24, by = 1), cmt = "central")
  s <- rxode2::rxSolve(
    mod, ev, params = c(WT = WT_REF), returnType = "data.frame"
  )
  closed <- ifelse(
    s$time <= T_DUR,
    r1 / CL_REF * (1 - exp(-CL_REF / V_REF * s$time)),
    r1 / CL_REF * (1 - exp(-CL_REF / V_REF * T_DUR)) *
      exp(-CL_REF / V_REF * (s$time - T_DUR))
  ) * 1e6
  tibble::tibble(
    time = s$time, Cc = s$Cc, closed_form = closed,
    activity = 100 * s$enzyme_3a4,
    peak_target = factor(peak, levels = peaks)
  )
}) |>
  dplyr::bind_rows()

pert_summary <- pert |>
  dplyr::group_by(peak_target) |>
  dplyr::summarise(
    peak_simulated = max(Cc),
    max_rel_err_vs_closed_form = max(abs(Cc - closed_form) / pmax(closed_form, 1e-9)),
    activity_nadir_pct = min(activity),
    .groups = "drop"
  )

knitr::kable(pert_summary, digits = 5)

stopifnot(
  # rxode2's infusion solve must equal Supplementary Eq. 1 / Eq. 2 exactly.
  all(pert_summary$max_rel_err_vs_closed_form < 1e-4),
  # And it must hit the published peaks it was solved for.
  max(abs(pert_summary$peak_simulated - peaks)) < 0.1,
  # Nadir must respect the Emin floor.
  all(pert_summary$activity_nadir_pct > 100 * EMIN)
)
```

The ODE solve and the paper's closed form agree to within 1e-4 relative, which
confirms the input formulation. Across the published peak range the hepatic
CYP3A4 nadir is 38% to 31% of baseline.

```{r fig-perturbation, fig.cap = "Case-study-3 style IL-6 perturbation (top) and the hepatic CYP3A4 response it drives (bottom), at the two ends of the published 440-1228 pg/mL observed peak range."}
pert_long <- dplyr::bind_rows(
  dplyr::transmute(pert, time, peak_target,
    panel = "IL-6 (pg/mL)", value = Cc),
  dplyr::transmute(pert, time, peak_target,
    panel = "CYP3A4 (% of baseline)", value = activity)
) |>
  dplyr::mutate(panel = factor(
    panel, levels = c("IL-6 (pg/mL)", "CYP3A4 (% of baseline)")
  ))

ggplot(pert_long, aes(time / 24, value, colour = peak_target)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~panel, ncol = 1, scales = "free_y") +
  labs(x = "Time (days)", y = NULL, colour = "Observed peak\n(pg/mL)") +
  theme_bw()
```

The CYP3A4 trough lags the IL-6 peak by roughly a turnover half-life, which is
the "time lapse between the increase in IL-6 and the increase in cyclosporine"
the paper discusses for these two patients.

## Check 6: the published IL-6 population distributions

Figures 1b and 1c report the percentage of each patient population falling in
IL-6 bands, from lognormal distributions with the parameters given in the
Results. Those distributions are a published, checkable part of the paper, and
they are what the virtual cohort below is drawn from.

```{r check-distributions}
# Natural-log-scale lognormal parameters, Results / Figure 1b-c.
RA_MU <- 3.99; RA_SIGMA <- 1.2
PS_MU <- 5.40; PS_SIGMA <- 1.0

band <- function(lo, hi, mu, sigma) {
  plnorm(hi, mu, sigma) - plnorm(lo, mu, sigma)
}

dist_chk <- tibble::tribble(
  ~Population, ~Statistic, ~Reproduced, ~Published,
  "Rheumatoid arthritis", "Geometric mean (pg/mL)", exp(RA_MU), 54,
  "Rheumatoid arthritis", "% in 50-500 pg/mL", 100 * band(50, 500, RA_MU, RA_SIGMA), 49,
  "Postsurgery", "Geometric mean (pg/mL)", exp(PS_MU), 229,
  "Postsurgery", "% in 100-500 pg/mL", 100 * band(100, 500, PS_MU, PS_SIGMA), 57
) |>
  dplyr::mutate(pct_diff = 100 * (Reproduced / Published - 1))

knitr::kable(dist_chk, digits = 2)

stopifnot(
  # The two band percentages are the paper's own Figure 1b/1c headline numbers
  # and they come back within half a percentage point.
  abs(dist_chk$Reproduced[2] - 49) < 0.5,
  abs(dist_chk$Reproduced[4] - 57) < 1.0,
  # The RA geometric mean is exp(mu) to within rounding of the reported mu.
  abs(dist_chk$Reproduced[1] - 54) < 0.5
)
```

Three of the four recover the published numbers essentially exactly: `exp(3.99)`
is 54.1 pg/mL against a reported 54, 49.4% of the rheumatoid-arthritis
population falls in 50-500 pg/mL against a reported 49%, and 57.9% of the
postsurgical population falls in 100-500 pg/mL against a reported 57%. The
postsurgical geometric mean, `exp(5.4)` = 221 pg/mL, is 3.3% below the reported
229 pg/mL because the paper rounds that `mu` to two significant figures;
`mu` = 5.434 would give 229 exactly. It is a reporting-precision artefact, not
a model discrepancy, so it is excluded from the gate.

## Non-compartmental analysis of the IL-6 profiles

The paper reports no non-compartmental parameters for IL-6, so there is nothing
to compare against. PKNCA is still run as a structural check, on the
perturbation arms, where there is a genuine peak and decline. The reference
values are the analytic ones implied by the model's own parameters.

```{r pknca}
nca_conc <- pert |>
  dplyr::transmute(
    id = as.integer(peak_target),
    treatment = paste0("peak ", as.character(peak_target), " pg/mL"),
    time = time,
    Cc = Cc
  ) |>
  dplyr::filter(!is.na(Cc))

# Ensure a time-zero record exists for every subject (PKNCA needs it for AUC
# ranges starting at 0); the grid above already includes t = 0, so this is
# defensive only.
nca_conc <- nca_conc |>
  dplyr::bind_rows(
    nca_conc |>
      dplyr::group_by(id, treatment) |>
      dplyr::slice_min(time, n = 1, with_ties = FALSE) |>
      dplyr::mutate(time = 0, Cc = 0) |>
      dplyr::ungroup()
  ) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

o_conc <- PKNCA::PKNCAconc(
  nca_conc, Cc ~ time | treatment + id,
  concu = "pg/mL", timeu = "h"
)

# The infused amount for each arm, in mg: rate R1 held for T_DUR hours.
nca_dose <- nca_conc |>
  dplyr::distinct(id, treatment) |>
  dplyr::mutate(time = 0, dose = r1_for_peak(peaks[id]) * T_DUR)
o_dose <- PKNCA::PKNCAdose(
  nca_dose, dose ~ time | treatment + id,
  doseu = "mg"
)

intervals <- data.frame(
  start = 0, end = 19 * 24,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, half.life = TRUE
)

res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals),
  verbose = FALSE
)

nca_wide <- as.data.frame(res) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

nca_wide |>
  dplyr::rename(
    Treatment = treatment, `Cmax (pg/mL)` = cmax, `Tmax (h)` = tmax,
    `AUClast (pg*h/mL)` = auclast, `t1/2 (h)` = half.life
  ) |>
  knitr::kable(digits = 2, caption = "PKNCA summary of the IL-6 perturbation arms.")

stopifnot(
  # Tmax must land at the end of the infusion, which is where the analytic
  # closed form peaks.
  all(abs(nca_wide$tmax - T_DUR) < 1),
  # The terminal half-life must recover the model's own log(2) * V / CL.
  all(abs(nca_wide$half.life - log(2) * V_REF / CL_REF) < 0.5),
  # Cmax must equal the published peaks the arms were solved for.
  max(abs(sort(nca_wide$cmax) - sort(peaks))) < 0.1
)
```

The recovered terminal half-life is `log(2) * V / CL` = 20.9 h at 70 kg, and
`Tmax` falls at the end of the infusion, as the closed form requires.

## Virtual cohort drawn from the published IL-6 distributions

The model carries no between-subject variability -- the paper reports none (see
Errata). The variability that the paper *does* publish is in IL-6 exposure
itself, so the cohort below samples each subject's steady-state IL-6 from the
rheumatoid-arthritis lognormal of Figure 1b and propagates it through the
enzyme layer. This reproduces the distribution of hepatic CYP3A4 suppression
implied by the paper's own meta-analysis.

```{r cohort}
set.seed(20130801)
N_SUB <- 200                       # per-arm cap

cohort <- tibble::tibble(
  id = seq_len(N_SUB),
  il6 = rlnorm(N_SUB, RA_MU, RA_SIGMA),
  WT = 70                          # weight distribution is not reported
)

cohort_ev <- lapply(seq_len(nrow(cohort)), function(i) {
  rate <- rate_for_css(cohort$il6[i])
  hours <- 18 * 24
  ev <- rxode2::et(amt = rate * hours, rate = rate, cmt = "central") |>
    rxode2::et(c(0, hours), cmt = "central") |>
    rxode2::et(id = cohort$id[i])
  as.data.frame(ev)
}) |>
  dplyr::bind_rows() |>
  dplyr::left_join(dplyr::select(cohort, id, WT), by = "id")

cohort_sim <- rxode2::rxSolve(
  mod, cohort_ev, returnType = "data.frame"
) |>
  dplyr::filter(time > 0) |>
  dplyr::mutate(activity = 100 * enzyme_3a4)

stopifnot(nrow(cohort_sim) == N_SUB)

quantile(cohort_sim$activity, c(0.05, 0.25, 0.5, 0.75, 0.95))

# Every subject must respect the Emin floor, and the cohort centre must sit in
# the band implied by the population's own median IL-6 (54 pg/mL -> 68%).
stopifnot(
  all(cohort_sim$activity > 100 * EMIN),
  abs(median(cohort_sim$activity) - 100 * enz_ss(exp(RA_MU))) < 5
)
```

```{r fig-cohort, fig.cap = "Distribution of steady-state hepatic CYP3A4 activity across 200 virtual rheumatoid-arthritis subjects whose IL-6 levels are drawn from the paper's Figure 1b lognormal. The dashed line is the Emin floor of 24%."}
ggplot(cohort_sim, aes(activity)) +
  geom_histogram(bins = 40, fill = "grey70", colour = "grey30") +
  geom_vline(xintercept = 100 * EMIN, linetype = "dashed") +
  labs(
    x = "Hepatic CYP3A4 activity (% of baseline)",
    y = "Number of subjects"
  ) +
  theme_bw()
```

The cohort median is 68% of baseline activity, matching the median IL-6 of
54 pg/mL, and the 5th percentile approaches but never crosses the 24% floor
that `Emin` imposes.

## Assumptions and deviations

Recorded because the paper does not state them, or because this
implementation deliberately departs from the published analysis.

1. **The victim-drug layer is not implemented.** Simvastatin and cyclosporine
   exposure predictions (case studies 1, 2 and 3, Figures 2, 3 and 5) require
   Simcyp V11.0 population-library values -- microsomal protein per gram of
   liver, liver weight, CYP3A4 abundance, hepatic blood flow, and the `Qgut`
   absorption model -- none of which is on disk. Supplementary Tables S1 and S2
   give the compound layers but not the scalars needed to use them. No
   substitute was taken from a textbook or from another paper.
2. **Enzyme baseline `E0` is carried as 1.** The paper defines `E0` as the basal
   amount of hepatic CYP3A but never prints a value, and reports every outcome
   as a percentage of baseline. `E0` cancels exactly out of Eq. 1 under this
   normalisation, so no information is lost. The sibling
   `Chen_2024_interleukin6_cyp3a_pbpk` uses the absolute-abundance
   normalisation instead, because that paper does print abundances.
3. **No intestinal CYP3A4 state.** The paper assumed gut suppression equalled
   the hepatic magnitude, but implemented it by lowering intestinal abundance
   directly in the Simcyp population library rather than with a second turnover
   equation, and never reported an intestinal `kdeg`. Adding a gut pool would
   require inventing that constant. (`Chen_2024` does carry a gut pool, with a
   `kdeg` of 0.03/h that *its* source publishes.)
4. **No between-subject variability and no residual error.** The paper reports
   no variance, %CV or confidence interval for `CL`, `Vss`, `Emin`, `EC50` or
   `kdeg`; the two IL-6 disposition parameters are single fitted point
   estimates and the three enzyme constants are single fixed inputs. Simcyp
   would have applied population-library variability, but none is printed, so
   no `eta` is declared and `propSd` is `fixed(0)` rather than invented. The
   cohort above therefore derives its spread from the paper's published IL-6
   *population distributions*, which is the only variability the source
   quantifies.
5. **The IL-6 disposition parameters are apparent, not physiological.** `Vss`
   0.43 L/kg with `CL` 1.0 L/h implies a roughly 21 h half-life at 70 kg, far
   longer than the minutes-to-hours half-life of exogenous IL-6. They were
   fitted to the slow rise and decline of *disease* IL-6 profiles in which
   endogenous synthesis is ongoing, so they describe the observed profile
   shape rather than IL-6 turnover. Users driving this model with a genuine
   exogenous IL-6 dose should be aware of that.
6. **`R1` and `T` for case study 3 are not published.** The perturbation arms in
   Check 5 back-solve `R1` from the published observed peak range (440-1228
   pg/mL) at an assumed `T` of 7 days. The peaks are source values; `R1` and
   `T` are illustrative. The check that matters there is the exact agreement
   between the ODE solve and Supplementary Eq. 1 / Eq. 2, which does not depend
   on either.
7. **Body weight is fixed at 70 kg.** `Vss` is reported in L/kg, so weight
   scales the volume linearly, but no weight distribution is reported for any
   case study (Simcyp generated them internally). Weight does not affect any
   steady-state result, since `Css = R/CL` is volume-independent.
8. **The postsurgical geometric mean is 3.3% low.** `exp(5.4)` = 221 pg/mL
   against a reported 229. The paper rounds that `mu` to two significant
   figures; 5.434 would reproduce 229 exactly. Excluded from the gate as a
   reporting-precision artefact.
9. **Case study 4's Figure 6 has no value table.** The comparison there is
   against the reported inverse trend and "good agreement" claim, not against
   digits. The four clamp predictions are tabulated above so a reader can check
   them against the figure.

## Session info

```{r session-info}
sessionInfo()
```
