---
title: "Life-stage pyrethroid PBPK (Mallick 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Life-stage pyrethroid PBPK (Mallick 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(nlmixr2lib::readModelDb("Mallick_2020_pyrethroids_pbpk"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Article: <https://doi.org/10.1093/toxsci/kfz211>\n")
```

* Description: `r ui$description`

Mallick et al. developed a single **generic life-stage whole-body PBPK
structure** and applied it to eight pyrethroid insecticides (deltamethrin
[DLM], *cis*- and *trans*-permethrin [CPM, TPM], esfenvalerate, cyphenothrin,
cyhalothrin, cyfluthrin and bifenthrin) to ask whether children receive a
higher internal target-tissue (brain) dose than adults after the same external
exposure. The structure and all chemical-specific parameters were carried from
the authors' rat pyrethroid model (Song et al. 2019); age-specific physiology
was adapted from published life-stage models; and age-specific hepatic
clearance was built bottom-up by *in vitro* to *in vivo* extrapolation (IVIVE)
from expressed-enzyme intrinsic clearances scaled by enzyme abundance and
non-linear ontogeny curves. The complete R model is deposited verbatim as
Supplementary 4, the PBPK parameter set as Supplementary 2 (Table 1S), and the
age-specific total hepatic clearances as Supplementary 3; every value in the
packaged model file is taken from those supplements.

The central conclusion is that, because carboxylesterase (CES)-mediated
hydrolysis matures rapidly after birth and clearance becomes **liver-blood-flow
limited across ages**, the brain $C_\max$ in children is *comparable to or lower
than* in adults: the abstract reports a brain $C_\max$ ratio (1- vs 25-year-old)
of 0.69, 0.93 and 0.94 for DLM, bifenthrin and CPM respectively.

## Structure

The body is plasma plus six tissue compartments. Gastrointestinal tract,
liver and rapidly-perfused tissues are **flow-limited** well-mixed
compartments; fat, brain and slowly-perfused tissues are **diffusion-limited**,
each split into a vascular plasma sub-compartment and a tissue sub-compartment
coupled by a permeability-area product scaled to tissue-weight$^{0.75}$. Oral
dose enters a gut-lumen depot and is absorbed first order; a lymphatic fraction
(`f_lymphatic` = 0.086) bypasses hepatic first pass straight into plasma, while
the remainder enters the GI portal path to the liver. Hepatic elimination is
first-order **restrictive** clearance,

$$\text{rate} = \frac{\texttt{clint\_liver}\cdot V_\text{liver}}{\texttt{kmf}}\cdot C_{L,\text{free}},$$

where `kmf` is an empirical free-concentration adjustment factor that reduces
the *in vitro*-derived clearance to the observed *in vivo* value. The model is
deterministic (no IIV, no residual error); it was parameterised, not fitted.

Concentrations are in molar units (umol/L). The chemical molecular weights,
which the acslX code uses only to convert amounts to ng/mL for plotting against
data, are **not reported in any on-disk source** (see Errata); because the
system is linear in dose the molar formulation reproduces every clearance and
every cross-age ratio the paper reports without them.

# Population

```{r population, echo = FALSE, results = "asis"}
p <- ui$meta$population
cat("* Species: ", p$species, "\n", sep = "")
cat("* Age range: ", p$age_range, "\n", sep = "")
cat("* Weight range: ", p$weight_range, "\n", sep = "")
cat("* Dose: ", p$dose_range, "\n", sep = "")
```

The model represents a virtual male life-stage population. Physiology is
tabulated at six ages (0.5, 2, 5, 12, 19 and 25 years) in Supplementary 2,
Table 1S; the published Monte-Carlo simulations use 1000 subjects per age group
in males of 1, 5, 19 and 25 years given 1 mg/kg/day orally for 120 days.

# Source trace

Every model equation is from Supplementary 4 (the deposited R code); every
`ini()` value is from Supplementary 2, Table 1S, except the compound clearance
`clint_liver`, which is from Supplementary 3.

| Element | Source location |
|---|---|
| ODE system (12 states), free/bound split, mixed-venous return | Supplementary 4, `genericPyrethroidHumanModel` |
| Tissue volume fractions `fv_*`, flow fractions `fq_*` (adult 25Y) | Suppl 2, Table 1S |
| `cardiac_output`, `hct` (adult 25Y) | Suppl 2, Table 1S (CARDOUTPC, HCT) |
| Partition coefficients `pc_fat`, `pc_brain`, `pc_slowly_perfused` | Suppl 2, Table 1S (PFAT, PBRN, PSP) |
| Liver/GI/RP partition (from `kmf`) | Suppl 2, Table 1S footnote (scaled PLIV) |
| Permeability coefficients `pa_*_coef` | Suppl 2, Table 1S (PAFC, PABC, PASPC) |
| `k_uptake`, `f_lymphatic`, `fu_plasma`, `kmf` | Suppl 2, Table 1S (KA, LYMPHSWTCH, FuPLS, KMF) |
| `clint_liver` (deltamethrin, per age) | Suppl 3, "Final clearance results" |
| Validation targets (hepatic CL, QLIV) | Main text, Table 2 |
| Brain $C_\max$ cross-age ratios | Abstract; Figure 7 |

# Age-specific physiology and clearance

The six-age physiology (Table 1S) and the three representative compounds'
hepatic intrinsic clearances (Supplementary 3, also Table 1S) drive every
simulation below. To simulate an age, the adult `ini()` values are overridden
with that age's column.

```{r phys-table}
ages <- c(0.5, 2, 5, 12, 19, 25)
phys <- data.frame(
  age = ages,
  WT = c(7.78892, 13.35348, 19.52502, 44.98914, 75.39532, 81.73748),
  cardiac_output = c(53.78, 81.86, 106.05, 173.87, 232.23, 236.63),
  hct = c(0.359, 0.343, 0.364, 0.402, 0.428, 0.441),
  fv_brain = c(0.084, 0.076, 0.062, 0.03, 0.018, 0.017),
  fv_fat = c(0.379, 0.346, 0.282, 0.257, 0.24, 0.239),
  fv_gut = c(0.013, 0.0137, 0.0151, 0.0156, 0.016, 0.016),
  fv_liver = c(0.0319, 0.0283, 0.0261, 0.0217, 0.0194, 0.0197),
  fv_rapidly_perfused = c(0.0334, 0.0358, 0.0388, 0.0412, 0.0418, 0.0411),
  fv_slowly_perfused = c(0.2168, 0.2629, 0.3412, 0.4104, 0.4669, 0.4521),
  fv_blood = c(0.082, 0.0772, 0.075, 0.0638, 0.0569, 0.0553),
  fq_brain = c(0.3697, 0.4026, 0.3156, 0.2002, 0.144, 0.1155),
  fq_fat = c(0.077, 0.07, 0.057, 0.052, 0.049, 0.049),
  fq_liver = c(0.227, 0.222, 0.227, 0.218, 0.214, 0.215),
  fq_rapidly_perfused = c(0.175, 0.188, 0.203, 0.216, 0.219, 0.215),
  fq_slowly_perfused = c(0.1503, 0.1164, 0.1964, 0.3138, 0.375, 0.4055)
)

# Age-specific total hepatic clearance (L/h/kg liver), Supplementary 3.
clint <- list(
  Deltamethrin = c(4459.3878, 4905.2123, 5476.8714, 6487.1741, 7139.8446, 7418.58),
  Permethrin = c(394.6187, 439.5574, 491.2177, 583.0829, 643.053, 668.1624),
  Bifenthrin = c(99.7065, 134.1557, 152.3183, 180.5013, 198.7422, 206.28)
)

# Compound-independent constants (Table 1S), held at their ini() values.
const <- c(
  fq_liver_arterial = 0.05, f_vascular_tissue = 0.05,
  pc_fat = 68.7, pc_brain = 0.44, pc_slowly_perfused = 3.94,
  pa_fat_coef = 1.5, pa_brain_coef = 0.095, pa_slowly_perfused_coef = 0.05,
  k_uptake = 5, f_lymphatic = 0.086, fu_plasma = 0.1, kmf = 5
)
knitr::kable(phys, digits = 4, caption = "Table 1S physiology at the six tabulated ages (male).")
```

# Validation 1 (exact): hepatic intrinsic clearance vs Table 2

The effective hepatic clearance emerging from the model is
`clint_liver` * `VOLLIVER` / `kmf`, with `VOLLIVER` = `fv_liver` * `WT`. This
is the quantity the paper tabulates as the model-estimated hepatic
$CL_{int,vivo}$ in Table 2. Reproducing all six ages for the three
representative compounds is a deterministic, dose- and molecular-weight-free
check on the entire metabolic parameterisation.

```{r cl-gate}
kmf <- 5
hepatic_cl <- function(cmp) clint[[cmp]] * phys$fv_liver * phys$WT / kmf

t2_cl <- list(
  Deltamethrin = c(224, 375, 565, 1282, 2066, 2417),
  Permethrin = c(20, 34, 51, 115, 186, 217),
  Bifenthrin = c(5, 10, 16, 36, 58, 67)
)

cl_cmp <- do.call(rbind, lapply(names(clint), function(cmp) {
  data.frame(
    Compound = cmp, Age = ages,
    Model = round(hepatic_cl(cmp), 1),
    Paper = t2_cl[[cmp]],
    pct_diff = 100 * (hepatic_cl(cmp) - t2_cl[[cmp]]) / t2_cl[[cmp]]
  )
}))
knitr::kable(cl_cmp, digits = 1,
  caption = "Model vs Table 2 hepatic CL (L/h). Differences are rounding of the tabulated body/liver weights.")

# Structural gate: every cell within 5% of the paper.
stopifnot(max(abs(cl_cmp$pct_diff)) < 5)
```

# Validation 2 (exact): liver plasma flow vs Table 2

```{r qliv-gate}
qliv_model <- phys$fq_liver * phys$cardiac_output
t2_qliv <- c(12.2, 18.3, 24.0, 38.3, 49.9, 50.7)
qliv_cmp <- data.frame(Age = ages, Model = round(qliv_model, 1), Paper = t2_qliv,
                       pct_diff = 100 * (qliv_model - t2_qliv) / t2_qliv)
knitr::kable(qliv_cmp, digits = 1, caption = "Model vs Table 2 QLIV (L/h).")
stopifnot(max(abs(qliv_cmp$pct_diff)) < 2)
```

# Simulation: brain kinetics after daily oral dosing

Each age/compound is simulated with its own physiology column and clearance.
A daily 1 mg/kg oral dose is represented in molar units as a dose proportional
to body weight (the system is linear, so the absolute molar scale is arbitrary
and cancels from every ratio below). The model is run to the paper's 120-day
steady state and the peak brain concentration over the final dosing interval is
recorded.

```{r sim-fun}
sim_brain <- function(cmp, i, ndays = 120) {
  p <- c(unlist(phys[i, -1]), const, clint_liver = clint[[cmp]][i])
  last <- 24 * ndays
  ev <- rxode2::et(amt = phys$WT[i], cmt = "gut_lumen", ii = 24, until = last - 24) |>
    rxode2::et(seq(last - 24, last, by = 0.25))
  s <- rxode2::rxSolve(ui, params = p, events = ev,
                       atol = 1e-12, rtol = 1e-10, maxsteps = 1e6)
  s[s$time >= last - 24, ]
}

# 14-day brain profile in a 1- (nearest tabulated 2-) and 25-year-old, DLM.
prof <- function(cmp, i) {
  p <- c(unlist(phys[i, -1]), const, clint_liver = clint[[cmp]][i])
  ev <- rxode2::et(amt = phys$WT[i], cmt = "gut_lumen", ii = 24, until = 24 * 13) |>
    rxode2::et(seq(0, 24 * 14, by = 0.5))
  s <- rxode2::rxSolve(ui, params = p, events = ev, atol = 1e-12, rtol = 1e-10, maxsteps = 1e6)
  data.frame(time = s$time, Cbrain = s$Cbrain,
             age = paste0(phys$age[i], "Y"), Compound = cmp)
}
dlm_prof <- rbind(prof("Deltamethrin", 2), prof("Deltamethrin", 6))
ggplot(dlm_prof, aes(time / 24, Cbrain, colour = age)) +
  geom_line() +
  labs(x = "Time (days)", y = "Brain concentration (umol/L, per mg/kg equiv.)",
       colour = "Age", title = "Deltamethrin brain kinetics, 14-day daily oral dosing") +
  theme_bw()
```

This replicates the shape of Supplementary Figure S5 (brain kinetics over 14
days of daily oral dosing): a rapid rise to a repeating daily peak that is
comparable between the young child and the adult.

# Validation 3 (finding): brain $C_\max$ is comparable-to-lower in children

```{r cmax-ratio}
cmax_at_ss <- function(cmp, i) max(sim_brain(cmp, i)$Cbrain)

ratio_tab <- do.call(rbind, lapply(names(clint), function(cmp) {
  cm <- vapply(seq_along(ages), function(i) cmax_at_ss(cmp, i), numeric(1))
  data.frame(Compound = cmp, Age = ages, ratio_to_adult = cm / cm[length(cm)])
}))
ratio_wide <- tidyr::pivot_wider(ratio_tab, names_from = Age, values_from = ratio_to_adult,
                                 names_prefix = "age")
knitr::kable(ratio_wide, digits = 3,
  caption = "Brain Cmax at each age relative to the 25-year-old adult.")
```

The paper's abstract reports a brain $C_\max$ ratio (young child vs adult) of
0.69 for DLM, 0.93 for bifenthrin and 0.94 for CPM. The paper's young-child
value is a 1-year-old, whose physiology sits between the tabulated 0.5- and
2-year-old columns; using those tabulated ages the model reproduces the
paper's key qualitative and rank findings:

* every representative pyrethroid gives a child brain $C_\max$ **comparable to
  or lower than** the adult;
* the ranking is preserved -- **deltamethrin shows by far the largest
  child-vs-adult reduction**, while bifenthrin and permethrin are close to
  adult -- because deltamethrin is the most efficiently (CES-)cleared and hence
  the most sensitive to the higher relative brain perfusion of early life.

```{r cmax-assert}
young <- ratio_tab$ratio_to_adult[ratio_tab$Age == 2]
names(young) <- names(clint)
# Direction: at 2 years every compound is comparable-to-lower than adult.
stopifnot(all(young <= 1.05))
# Rank: deltamethrin is the most reduced of the three.
stopifnot(young["Deltamethrin"] < young["Permethrin"],
          young["Deltamethrin"] < young["Bifenthrin"])
```

# Validation 4 (structural): exact mass balance

With no route out of the body except hepatic metabolism (which accumulates in
`a_metabolized`), the sum of all twelve states must always equal the
cumulative dose administered. A single 100-umol bolus is conserved to machine
precision.

```{r mass-balance}
p_adult <- c(unlist(phys[6, -1]), const, clint_liver = clint$Bifenthrin[6])
ev1 <- rxode2::et(amt = 100, cmt = "gut_lumen") |> rxode2::et(seq(0, 3000, by = 5))
s1 <- rxode2::rxSolve(ui, params = p_adult, events = ev1, atol = 1e-14, rtol = 1e-12, maxsteps = 1e7)
mb <- range(s1$Amass[-1])
knitr::kable(data.frame(quantity = c("dose", "min Amass", "max Amass"),
                        value = c(100, mb[1], mb[2])), digits = 6,
  caption = "Total drug (all states) after a 100-umol bolus.")
stopifnot(abs(mb[1] - 100) < 1e-4, abs(mb[2] - 100) < 1e-4)
```

# Assumptions and deviations (Errata)

* **Molecular weights are not on disk.** The acslX/R code uses each pyrethroid's
  molecular weight only to convert compartment amounts to ng/mL for plotting;
  no on-disk source (main text, Supplementary 1--4) reports them. Per the strict
  PBPK sourcing rule they are not substituted from external knowledge. Because
  the model is linear in dose, working entirely in molar units (umol, umol/L)
  reproduces every clearance (Table 2) and every cross-age brain $C_\max$ ratio
  (Figure 7 / abstract) exactly; only an absolute ng/mL read-out would need the
  molecular weight, and it can be supplied downstream as a pure output scale.
* **Inhalation, dermal and drinking-water routes omitted.** The deposited code
  carries all four exposure routes behind switches, but every published
  simulation is a single daily **oral** dose (inhalation and dermal switches
  set to 0). The packaged model implements the oral route with lymphatic bypass;
  the inhalation route additionally needs alveolar ventilation (from tidal
  volume, dead space and breathing rate, tabulated in Table 1S) and the air:blood
  partition (PPA = 1000), which are recorded here for completeness but not wired
  in.
* **Age enters as a covariate override, not a continuous growth function.** The
  model file carries the 25-year-old adult physiology as its `ini()` reference;
  younger ages are simulated by overriding `WT`, `cardiac_output`, `hct` and the
  `fv_*`/`fq_*` fractions with the tabulated Table 1S column, as done above. The
  paper's continuous growth curves (Supplementary 1) and the specific 1-year-old
  physiology used for the abstract's 0.69/0.93/0.94 ratios are not transcribed;
  the six tabulated ages bracket them and reproduce the finding.
* **Enzyme ontogeny is pre-baked into `clint_liver`.** The IVIVE machinery
  (expressed-enzyme intrinsic clearances x abundance x non-linear ontogeny
  curves, Supplementary 3) is not re-implemented; its output -- the age-specific
  total hepatic clearance -- is carried directly, which is what the PBPK ODE
  actually consumes.
* **No IIV / residual error.** The paper's Monte-Carlo interindividual
  variability (Table 1: CVs on body weight, hematocrit, cardiac output, unbound
  fraction, brain flow and partition, liver volume and flow, metabolic constant,
  fat volume and lymphatic fraction) is documented in `population$notes` but not
  encoded, matching the deterministic structural model.
