---
title: "Gemcabene, statins and ezetimibe LDL-C MBMA (Mandema 2005)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gemcabene, statins and ezetimibe LDL-C MBMA (Mandema 2005)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7.5, fig.height = 5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Mandema_2005_gemcabene_mbma"))$reference`
* Description: `r rxode2::rxode(readModelDb("Mandema_2005_gemcabene_mbma"))$description`
* Article: <https://doi.org/10.1208/aapsj070352> (PMC2751254, open access)

This is a model-based meta-analysis (MBMA) built to support a go/no-go decision
on gemcabene, an investigational lipid-altering agent (development code
CI-1027). The analysis pooled published statin and ezetimibe trials with four
Pfizer-sponsored gemcabene trials so that gemcabene could be compared against
its competitors on a common dose-response scale, including in combination with
a statin. On the strength of this model the sponsor stopped development: the
model showed that gemcabene's benefit *on top of* a statin collapses at high
statin doses, unlike ezetimibe's.

# Population

Mandema 2005 fitted study-arm summary means from 25 randomized controlled
trials totalling 9,886 patients (the paper's Abstract rounds this to "almost
10,000"). Twenty-one trials evaluated statins, ezetimibe, or their combination
in patients with hypercholesterolemia; all were randomized, double-blind,
multiple-dose once-daily parallel-group studies, 13 of them placebo controlled,
and most multicentre. They were identified by a Medline search with a 31 May
2003 cut-off (minimum 4 weeks of treatment), supplemented by FDA
summary-basis-of-approval documents.

Four further trials came from the gemcabene programme. Three studied gemcabene
monotherapy in healthy volunteers, nondiabetic healthy obese subjects, and
subjects with low HDL-C and normal or elevated triglycerides; the fourth was
the phase IIA hypercholesterolemia trial that crossed gemcabene 300/600/900
mg/day with atorvastatin 10/20/80 mg/day. Three of the four were unpublished at
the time of writing.

Daily doses spanned atorvastatin 2.5-80 mg, rosuvastatin 1-80 mg, simvastatin
10-80 mg, lovastatin 10-80 mg, pravastatin 10-40 mg, ezetimibe 0.25-40 mg, and
gemcabene 50-900 mg (Table 1). Baseline LDL-C ranged 112-217 mg/dL across
trials. Critically, the three gemcabene healthy-subject trials had much *lower*
baseline LDL-C (112-120 mg/dL) than the hypercholesterolemia trials (165-217
mg/dL). The paper handled that mismatch upstream rather than in the
dose-response model: a per-trial ANOVA over those three trials (factors dose,
trial, baseline LDL-C, triglycerides) found "a small impact of baseline LDL-C
values < 100 mg/dL" and *no* treatment-by-baseline interaction, and its
least-squares means were then used as the summary data fed to the dose-response
analysis.

Each modelled observation is therefore a study-arm mean percentage change in
LDL-C from pretreatment baseline, not an individual patient value. The
simulation scope of this model is the study-arm mean; it is **not** suitable
for individual-subject simulation, and it has no PK layer and no time course.

The same information is available programmatically via
`rxode2::rxode(readModelDb("Mandema_2005_gemcabene_mbma"))$population`.

# Source trace

The model is Equations 1 and 2 of the paper (Materials and Methods, Statistical
Analysis). Equation 1 combines a statin with a nonstatin:

$$
Y = E_0 + E_{\text{statin}} + E_{\text{nonstatin}}
  + 0.01 \cdot g \cdot E_{\text{statin}} \cdot E_{\text{nonstatin}}
  + \eta + \varepsilon
$$

and Equation 2 is a sigmoid Emax dose-response applied to each drug:

$$
E_{\text{drug}} = \frac{\text{Dose}^{n} \cdot E_{\max}}{\text{Dose}^{n} + \text{ED}_{50}^{n}}
$$

$Y$ is the percentage change in LDL-C from pretreatment baseline, so every $E$
term is on the **percent** scale and a negative value is a reduction. The
factor $0.01$ is exactly what converts a product of two percentages back into a
percentage: at $g = 1$ Equation 1 reproduces pharmacological independence,
$1 - (1 - f_{\text{statin}})(1 - f_{\text{nonstatin}})$ on the fractional
scale. $g > 1$ means *less* than independent.

| Equation / parameter | Value | Source location |
|----------------------|------:|-----------------|
| Combination structure (Eq 1) | n/a | Methods, Statistical Analysis, p. E514 |
| Sigmoid Emax dose-response (Eq 2) | n/a | Methods, Statistical Analysis, p. E514 |
| `e0` (placebo effect, North American trials) | 0.802 % | Table 2, E0 [0.0598 to 1.54] |
| `e_region_europe_e0` (extra European placebo reduction) | -4 % | Results, LDL-C section (prose only; no Table 2 row, no CI) |
| `emax_statin` (shared across all five statins) | -78.7 % | Table 2, Emax, statin (%) [-90.7 to -66.7] |
| `led50_atorvastatin` | 13.1 mg/day | Table 2, ED50, Atorvastatin (mg) [6.57 to 26.2] |
| `led50_rsv` | 4.35 mg/day | Table 2, ED50, Rosuvastatin (mg) [2.19 to 8.62] |
| `led50_smv` | 30.5 mg/day | Table 2, ED50, Simvastatin (mg) [15 to 62.1] |
| `led50_lov` | 82.8 mg/day | Table 2, ED50, Lovastatin (mg) [37.1 to 185] |
| `led50_prv` | 97.3 mg/day | Table 2, ED50, Pravastatin (mg) [42.4 to 223] |
| `ln_statin` (Hill coefficient, shared across statins) | 0.451 | Table 2, n statin [0.366 to 0.557] |
| `emax_ezt` | -19.6 % | Table 2, Emax,Ezetimibe (%) [-20.6 to -18.6] |
| `led50_ezt` | 0.302 mg/day | Table 2, ED50,Ezetimibe (mg) [0.151 to 0.604] |
| `ln_ezt` (ezetimibe Hill coefficient; held constant) | 1 | Table 2, n Ezetimibe (no CI); Results: "A Hill coefficient for ezetimibe could not be estimated and was fixed to 1." |
| `gamma_ezt` (statin x ezetimibe interaction; held constant) | 1 | Table 2, g Ezetimibe (no CI); Results: "not statistically significantly different from 1 and was, therefore, fixed to this pharmacological value" |
| `emax_gem` | -34.8 % | Table 2, Emax,gemcabene (%) [-45 to -24.6] |
| `led50_gem` | 314 mg/day | Table 2, ED50,gemcabene (mg) [220 to 448] |
| `ln_gem` (gemcabene Hill coefficient) | 2.27 | Table 2, n gemcabene [1.19 to 4.34] |
| `gamma_gem` (statin x gemcabene interaction) | 1.69 | Table 2, g gemcabene [1.49 to 1.88]; Results also give 1.69 +/- 0.10 (mean +/- SE), p < 0.001 versus 1 |
| Trial random-effect variance (nu^2, Eq 1) | not encoded | Results: "The variance of the trial-specific random effect was found to be not statistically significantly different from zero"; no numeric estimate published |
| Residual variance (sigma^2, Eq 1) | `addSd` held at 0 | Eq 1 declares eps ~ N(0, sigma^2); **no numeric value for sigma is published anywhere in the paper** |

## A note on the published PDF's minus signs

The AAPS Journal PDF renders the minus sign of every negative Table 2 entry as
a leading digit `2` (so `Emax, statin` prints as `278.7` for -78.7, and
`Emax,gemcabene` as `234.8` for -34.8). The same substitution turns
`g = 1.69 +/- 0.10` into `g 5 1.69 6 0.10`. The signs used in the model file
are not a guess: they are confirmed end-to-end by reproducing every cell of the
paper's own Table 3 from Table 2, which is done below.

# Errata

No published erratum, corrigendum, or author correction was located for
Mandema 2005. Searches of the AAPS Journal / PMC landing page for
PMC2751254, the DOI 10.1208/aapsj070352, and PubMed (PMID 16353928) returned no
correction notices as of the extraction date (2026-09-01).

Reporting gaps in the paper as published (not errata, but they constrain what
can be encoded):

* No numeric estimate of the residual SD `sigma` is given anywhere, although
  Equation 1 declares it.
* The 4% larger placebo response in European trials appears only as a prose
  sentence, to one significant figure, with no Table 2 row and no confidence
  interval.
* Table 2 reports marginal 95% confidence intervals but not the
  variance-covariance matrix, so the paper's own uncertainty bands (derived
  from 5,000 parameter draws) cannot be reproduced. See Assumptions below.

# Virtual cohort

There is no cohort to simulate in the usual sense: the model is a
deterministic, time-independent, per-arm dose-response with no between-subject
variability, no trial random effect (its variance was not different from zero),
and a residual SD held at zero. Every check below is therefore a *deterministic*
comparison against a published number, and the assertions are correspondingly
tight -- there is no simulated cohort whose extreme could shift between rxode2
versions or thread counts.

Each "subject" in the event tables below is one study arm: a set of per-drug
daily doses plus the trial's region indicator.

```{r virtual-design}
mod <- readModelDb("Mandema_2005_gemcabene_mbma")

# Every covariate column the model consumes, defaulted to a North American
# placebo arm. Helper `arm()` overrides only what a given arm actually gets.
arm_template <- list(
  CONMED_ATORVASTATIN_DOSE = 0,
  CONMED_RSV_DOSE          = 0,
  CONMED_SMV_DOSE          = 0,
  CONMED_LOV_DOSE          = 0,
  CONMED_PRV_DOSE          = 0,
  CONMED_EZT_DOSE          = 0,
  CONMED_GEMCABENE_DOSE    = 0,
  REGION_EUROPE            = 0
)

arm <- function(...) {
  over <- list(...)
  stopifnot(all(names(over) %in% names(arm_template)))
  as.data.frame(utils::modifyList(arm_template, over))
}

# Solve a data.frame of arms and return it with the predicted percentage
# change in LDL-C attached. The model is algebraic, so each arm is a single
# time-0 observation record; `keep` carries the labelling columns through.
solve_arms <- function(arms, keep = character()) {
  ev <- arms
  ev$id   <- seq_len(nrow(ev))
  ev$time <- 0
  ev$amt  <- 0
  ev$evid <- 0L
  # `mod` is eta-free, so `omega = NA` is unnecessary and errors on the
  # released rxode2 (5.1.6: "invalid 'times' argument").
  out <- as.data.frame(rxode2::rxSolve(mod, events = ev))
  stopifnot(nrow(out) == nrow(arms), !anyNA(out$Cc))
  cbind(arms[keep], pct_change_ldlc = out$Cc)
}
```

# Replication: statin monotherapy dose-response (Figure 1)

Figure 1 of the paper plots the model-predicted mean percentage LDL-C reduction
against dose for each of the five statins. The defining structural finding is
that the statins share one Emax and one (shallow) Hill coefficient and differ
only in ED50: "none of the statins had a statistically significant effect on
Emax when tested independently ... a similar maximal effect of about 79%
reduction in LDL-C over placebo."

```{r figure-1, fig.cap = "Replicates Figure 1 of Mandema 2005: model-predicted percentage change in LDL-C for monotherapy with each of the five statins. The five curves are the same shape shifted along the log-dose axis; only ED50 differs."}
statin_cols <- c(
  Atorvastatin = "CONMED_ATORVASTATIN_DOSE",
  Rosuvastatin = "CONMED_RSV_DOSE",
  Simvastatin  = "CONMED_SMV_DOSE",
  Lovastatin   = "CONMED_LOV_DOSE",
  Pravastatin  = "CONMED_PRV_DOSE"
)

dose_grid <- c(0.5, 1, 2.5, 5, 10, 20, 40, 80, 160)

statin_arms <- dplyr::bind_rows(lapply(names(statin_cols), function(sn) {
  a <- dplyr::bind_rows(lapply(dose_grid, function(d) {
    x <- list(d); names(x) <- statin_cols[[sn]]
    do.call(arm, x)
  }))
  a$statin <- sn
  a$dose   <- dose_grid
  a
}))

statin_curves <- solve_arms(statin_arms, keep = c("statin", "dose"))

ggplot(statin_curves, aes(x = dose, y = pct_change_ldlc, colour = statin)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.3) +
  scale_x_log10(breaks = c(0.5, 1, 2.5, 5, 10, 20, 40, 80, 160)) +
  labs(
    x      = "Daily statin dose (mg/day, log scale)",
    y      = "LDL-C % change from pretreatment",
    colour = NULL,
    title  = "Mandema 2005 Figure 1 -- statin monotherapy dose-response",
    caption = "Shared Emax (-78.7%) and Hill coefficient (0.451); statins differ only in ED50."
  ) +
  theme_bw() +
  theme(legend.position = "top")
```

## Shared-Emax and relative-potency checks

The Discussion states the relative potencies of rosuvastatin, simvastatin,
lovastatin and pravastatin against atorvastatin as 0.33, 2.3, 6.3 and 7.4, and
gives the doses equivalent to 10 mg atorvastatin as 3.3, 23, 63 and 74 mg. Both
are pure ED50 ratios under the shared-Emax structure, so they test whether the
five ED50 values were transcribed correctly *and* whether the shared-Emax
structure was implemented as the authors described it.

```{r relative-potency}
ed50 <- c(Atorvastatin = 13.1, Rosuvastatin = 4.35, Simvastatin = 30.5,
          Lovastatin = 82.8, Pravastatin = 97.3)
others <- c("Rosuvastatin", "Simvastatin", "Lovastatin", "Pravastatin")

potency <- data.frame(
  Statin            = others,
  potency_model     = round(unname(ed50[others] / ed50[["Atorvastatin"]]), 2),
  potency_paper     = c(0.33, 2.3, 6.3, 7.4),
  equiv_dose_model  = round(10 * unname(ed50[others] / ed50[["Atorvastatin"]]), 1),
  equiv_dose_paper  = c(3.3, 23, 63, 74)
)

knitr::kable(
  potency |>
    dplyr::rename(
      "Relative potency (model)"      = potency_model,
      "Relative potency (paper)"      = potency_paper,
      "Dose equal to atorvastatin 10 mg, model (mg)" = equiv_dose_model,
      "Dose equal to atorvastatin 10 mg, paper (mg)" = equiv_dose_paper
    ),
  caption = "Relative statin potencies against atorvastatin (Mandema 2005 Discussion). Deterministic ED50 ratios."
)

# Both sides are the same published ED50 values; the only difference is the
# paper's rounding to two significant figures, so these bounds are tight.
stopifnot(
  max(abs(potency$potency_model - potency$potency_paper)) < 0.05,
  max(abs(potency$equiv_dose_model - potency$equiv_dose_paper) /
        potency$equiv_dose_paper) < 0.02
)

# Emax check: at an effectively infinite statin dose the model must return the
# shared Emax of -78.7% (paper Results: "about 79% reduction ... over placebo").
# The probe dose has to be absurdly large precisely because the statin Hill
# coefficient is 0.451: the approach to Emax goes as (ED50/Dose)^0.451, so
# 1e9 mg is still 0.02 percentage points short while 1e15 mg is within 1e-3.
# That slow approach IS the paper's "shallow dose-response relationship ...
# the maximum efficacy is more difficult to achieve" (Discussion).
emax_chk <- solve_arms(arm(CONMED_ATORVASTATIN_DOSE = 1e15))$pct_change_ldlc -
  solve_arms(arm())$pct_change_ldlc
cat(sprintf("Statin effect at infinite dose: %.2f%% (Table 2 Emax,statin = -78.7%%)\n",
            emax_chk))
stopifnot(abs(emax_chk - (-78.7)) < 0.01)
```

# Replication: ezetimibe and gemcabene monotherapy (Figures 2 and 4)

Figure 2 shows ezetimibe monotherapy and Figure 4 shows gemcabene monotherapy.
The two nonstatins behave very differently: ezetimibe has a low Emax (-19.6%)
but an ED50 of only about 0.3 mg, so the marketed 10 mg dose sits essentially
*at* its maximum; gemcabene has a larger Emax (-34.8%) but an ED50 of 314 mg
and a steep Hill coefficient of 2.27, so 900 mg is still short of its maximum.

```{r figure-2-4, fig.cap = "Replicates Figures 2 and 4 of Mandema 2005: monotherapy dose-response for ezetimibe (ED50 0.302 mg, Hill 1) and gemcabene (ED50 314 mg, Hill 2.27). Points mark the doses each drug was actually studied at (Table 1)."}
ezt_grid <- c(0.05, 0.1, 0.25, 0.5, 1, 2.5, 5, 10, 20, 40)
gem_grid <- c(25, 50, 100, 150, 300, 450, 600, 900, 1800)

nonstatin_curves <- dplyr::bind_rows(
  solve_arms(
    dplyr::bind_rows(lapply(ezt_grid, function(d) arm(CONMED_EZT_DOSE = d))) |>
      dplyr::mutate(drug = "Ezetimibe", dose = ezt_grid),
    keep = c("drug", "dose")
  ),
  solve_arms(
    dplyr::bind_rows(lapply(gem_grid, function(d) arm(CONMED_GEMCABENE_DOSE = d))) |>
      dplyr::mutate(drug = "Gemcabene", dose = gem_grid),
    keep = c("drug", "dose")
  )
)

studied <- dplyr::bind_rows(
  data.frame(drug = "Ezetimibe", dose = c(0.25, 1, 5, 10, 20, 40)),
  data.frame(drug = "Gemcabene", dose = c(50, 150, 300, 450, 600, 900))
) |>
  dplyr::inner_join(nonstatin_curves, by = c("drug", "dose"))

ggplot(nonstatin_curves, aes(x = dose, y = pct_change_ldlc)) +
  geom_line(linewidth = 0.9, colour = "steelblue") +
  geom_point(data = studied, size = 2, colour = "black") +
  facet_wrap(~ drug, scales = "free_x") +
  scale_x_log10() +
  labs(
    x = "Daily dose (mg/day, log scale)",
    y = "LDL-C % change from pretreatment",
    title = "Mandema 2005 Figures 2 and 4 -- nonstatin monotherapy dose-response",
    caption = "Black points mark the dose strengths actually studied (Table 1)."
  ) +
  theme_bw()
```

The Discussion gives two point predictions for monotherapy, which are the
strictest available single-drug checks:

```{r monotherapy-points}
placebo <- solve_arms(arm())$pct_change_ldlc

mono <- data.frame(
  Treatment = c("Gemcabene 900 mg", "Ezetimibe 10 mg"),
  model = c(
    solve_arms(arm(CONMED_GEMCABENE_DOSE = 900))$pct_change_ldlc - placebo,
    solve_arms(arm(CONMED_EZT_DOSE = 10))$pct_change_ldlc - placebo
  ),
  paper = c(-31.9, -19.1)
)
mono$model <- round(mono$model, 2)

knitr::kable(
  mono |>
    dplyr::rename("Model (% change)" = model, "Paper (% change)" = paper),
  caption = "Monotherapy predictions quoted in the Mandema 2005 Discussion. Deterministic; differences are the paper's rounding of Table 2 inputs to three significant figures."
)

stopifnot(max(abs(mono$model - mono$paper)) < 0.15)
```

# Replication: the interaction between statins and the two nonstatins (Figures 3, 5, 6 and Table 3)

This is the heart of the paper, and the reason gemcabene was stopped. Table 3
tabulates the *additional* mean LDL-C change from adding 900 mg gemcabene or
10 mg ezetimibe on top of atorvastatin monotherapy at 0, 10, 20, 40 and 80 mg.
Because Table 3 reports within-trial differences, the intercept `e0` (and any
trial-level region effect) cancels exactly.

Table 3 is an independent output: it is not an input to the model, and every
cell in it is a nonlinear function of all of Table 2. Reproducing all ten mean
cells is therefore a genuine end-to-end confirmation of the structural form,
the parameter transcription, and every sign.

```{r table-3}
ator_doses <- c(0, 10, 20, 40, 80)

mono_ator <- solve_arms(
  dplyr::bind_rows(lapply(ator_doses, function(d) arm(CONMED_ATORVASTATIN_DOSE = d)))
)$pct_change_ldlc

add_gem <- solve_arms(
  dplyr::bind_rows(lapply(ator_doses, function(d) {
    arm(CONMED_ATORVASTATIN_DOSE = d, CONMED_GEMCABENE_DOSE = 900)
  }))
)$pct_change_ldlc

add_ezt <- solve_arms(
  dplyr::bind_rows(lapply(ator_doses, function(d) {
    arm(CONMED_ATORVASTATIN_DOSE = d, CONMED_EZT_DOSE = 10)
  }))
)$pct_change_ldlc

table3 <- data.frame(
  atorvastatin_mg = ator_doses,
  gem_model = round(add_gem - mono_ator, 1),
  gem_paper = c(-31.9, -12.0, -8.7, -5.5, -2.5),
  ezt_model = round(add_ezt - mono_ator, 1),
  ezt_paper = c(-19.1, -12.0, -10.8, -9.7, -8.7)
)

knitr::kable(
  table3 |>
    dplyr::rename(
      "Atorvastatin (mg)"         = atorvastatin_mg,
      "+900 mg gemcabene (model)" = gem_model,
      "+900 mg gemcabene (paper)" = gem_paper,
      "+10 mg ezetimibe (model)"  = ezt_model,
      "+10 mg ezetimibe (paper)"  = ezt_paper
    ),
  caption = "Replicates Mandema 2005 Table 3: additional mean % change in LDL-C from adding 900 mg gemcabene or 10 mg ezetimibe to atorvastatin monotherapy. Paper columns are the 'Mean' entries of Table 3."
)

t3_err <- c(abs(table3$gem_model - table3$gem_paper),
            abs(table3$ezt_model - table3$ezt_paper))
cat(sprintf("Max absolute deviation across all 10 Table 3 mean cells: %.2f percentage points\n",
            max(t3_err)))

# Deterministic: both sides are the published Table 2 parameters evaluated
# through the published Equations 1-2. The only source of difference is that
# the paper rounded Table 2 to three significant figures before computing
# Table 3 and then rounded Table 3 to one decimal place. A transposed digit,
# a dropped minus sign, or a wrong interaction coefficient moves these cells
# by whole percentage points, so 0.15 still fails loudly.
stopifnot(max(t3_err) <= 0.15)
```

```{r figure-6, fig.cap = "Replicates Figure 6 of Mandema 2005: atorvastatin monotherapy versus atorvastatin combined with 900 mg gemcabene (left) or 10 mg ezetimibe (right). Ezetimibe adds a roughly constant increment across the whole atorvastatin dose range; gemcabene's increment collapses as the atorvastatin dose rises."}
fig6_grid <- c(0, 1, 2.5, 5, 10, 20, 40, 80)

fig6 <- dplyr::bind_rows(
  solve_arms(dplyr::bind_rows(lapply(fig6_grid, function(d)
    arm(CONMED_ATORVASTATIN_DOSE = d))) |>
      dplyr::mutate(panel = "+ 900 mg gemcabene", arm_lab = "Atorvastatin alone",
                    dose = fig6_grid),
    keep = c("panel", "arm_lab", "dose")),
  solve_arms(dplyr::bind_rows(lapply(fig6_grid, function(d)
    arm(CONMED_ATORVASTATIN_DOSE = d, CONMED_GEMCABENE_DOSE = 900))) |>
      dplyr::mutate(panel = "+ 900 mg gemcabene", arm_lab = "Combination",
                    dose = fig6_grid),
    keep = c("panel", "arm_lab", "dose")),
  solve_arms(dplyr::bind_rows(lapply(fig6_grid, function(d)
    arm(CONMED_ATORVASTATIN_DOSE = d))) |>
      dplyr::mutate(panel = "+ 10 mg ezetimibe", arm_lab = "Atorvastatin alone",
                    dose = fig6_grid),
    keep = c("panel", "arm_lab", "dose")),
  solve_arms(dplyr::bind_rows(lapply(fig6_grid, function(d)
    arm(CONMED_ATORVASTATIN_DOSE = d, CONMED_EZT_DOSE = 10))) |>
      dplyr::mutate(panel = "+ 10 mg ezetimibe", arm_lab = "Combination",
                    dose = fig6_grid),
    keep = c("panel", "arm_lab", "dose"))
)

ggplot(fig6, aes(x = dose, y = pct_change_ldlc, colour = arm_lab, linetype = arm_lab)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.3) +
  facet_wrap(~ panel) +
  scale_colour_manual(values = c("Atorvastatin alone" = "black",
                                 "Combination" = "firebrick")) +
  scale_linetype_manual(values = c("Atorvastatin alone" = "solid",
                                   "Combination" = "dashed")) +
  labs(
    x = "Daily atorvastatin dose (mg/day)", y = "LDL-C % change from pretreatment",
    colour = NULL, linetype = NULL,
    title = "Mandema 2005 Figure 6 -- atorvastatin alone versus combination",
    caption = "g = 1.69 for gemcabene (less than independent) versus g = 1 for ezetimibe (independent)."
  ) +
  theme_bw() +
  theme(legend.position = "top")
```

## The pharmacological-independence identity (ezetimibe, g = 1)

The paper defines independence explicitly: "the expected LDL-C reduction can be
calculated as `1 - (1 - statin fractional LDL-C reduction) * (1 - ezetimibe
fractional LDL-C reduction)`. An independent interaction is represented in the
dose-response model by an interaction coefficient g of 1."

That is an algebraic identity the implementation must satisfy exactly for
ezetimibe (whose `g` is held at 1), and must *not* satisfy for gemcabene
(`g = 1.69`). Checking both directions confirms both the `0.01` scaling factor
in Equation 1 and that the two interaction coefficients are not swapped.

```{r independence-identity}
grid <- c(0, 2.5, 5, 10, 20, 40, 80)

statin_only <- solve_arms(dplyr::bind_rows(lapply(grid, function(d)
  arm(CONMED_ATORVASTATIN_DOSE = d))))$pct_change_ldlc - placebo
ezt_only <- solve_arms(arm(CONMED_EZT_DOSE = 10))$pct_change_ldlc - placebo
gem_only <- solve_arms(arm(CONMED_GEMCABENE_DOSE = 900))$pct_change_ldlc - placebo

combo_ezt <- solve_arms(dplyr::bind_rows(lapply(grid, function(d)
  arm(CONMED_ATORVASTATIN_DOSE = d, CONMED_EZT_DOSE = 10))))$pct_change_ldlc - placebo
combo_gem <- solve_arms(dplyr::bind_rows(lapply(grid, function(d)
  arm(CONMED_ATORVASTATIN_DOSE = d, CONMED_GEMCABENE_DOSE = 900))))$pct_change_ldlc - placebo

# Independence prediction, expressed as a percentage change (negative).
independent <- function(a, b) -100 * (1 - (1 + a / 100) * (1 + b / 100))

ind_ezt <- independent(statin_only, ezt_only)
ind_gem <- independent(statin_only, gem_only)

cat(sprintf("Ezetimibe: max |model - independence identity| = %.10f %%-points\n",
            max(abs(combo_ezt - ind_ezt))))
cat(sprintf("Gemcabene: max |model - independence identity| = %.2f %%-points (should be LARGE)\n",
            max(abs(combo_gem - ind_gem))))

# Ezetimibe: exact algebraic identity, so machine precision is the right bound.
stopifnot(max(abs(combo_ezt - ind_ezt)) < 1e-8)
# Gemcabene: must be clearly sub-independent (g = 1.69, not 1). At 80 mg
# atorvastatin the model gives -2.5% additional versus -13.4% under
# independence, so this departure is not marginal.
stopifnot(max(abs(combo_gem - ind_gem)) > 5)
```

The ezetimibe combination reproduces the independence identity to machine
precision, and the gemcabene combination departs from it by more than 10
percentage points at the top of the atorvastatin dose range. That is the
paper's central finding, restated as an assertion.

## The decision the model supported

```{r decision-summary}
decision <- data.frame(
  Statement = c(
    "At atorvastatin 10 mg, 900 mg gemcabene and 10 mg ezetimibe add the same ~12%",
    "At atorvastatin 80 mg, gemcabene adds only ~2.5% versus ezetimibe's ~8.7%",
    "900 mg gemcabene alone is about as effective as 5 mg atorvastatin"
  ),
  Model = c(
    sprintf("gemcabene %.1f%% vs ezetimibe %.1f%%", table3$gem_model[2], table3$ezt_model[2]),
    sprintf("gemcabene %.1f%% vs ezetimibe %.1f%%", table3$gem_model[5], table3$ezt_model[5]),
    sprintf("gemcabene 900 mg %.1f%% vs atorvastatin 5 mg %.1f%%",
            gem_only,
            solve_arms(arm(CONMED_ATORVASTATIN_DOSE = 5))$pct_change_ldlc - placebo)
  )
)
knitr::kable(decision, caption = "The three quantitative claims in the Mandema 2005 Discussion that drove the decision to stop gemcabene development.")

# "about the same LDL-C reduction as 5 mg of atorvastatin" (Discussion).
ator5 <- solve_arms(arm(CONMED_ATORVASTATIN_DOSE = 5))$pct_change_ldlc - placebo
stopifnot(abs(gem_only - ator5) < 1.5)
```

# The European placebo shift

The only retained covariate is trial location. Because it acts on the intercept
`E0`, it shifts every arm of a European trial equally and therefore cancels out
of any within-trial contrast such as Table 3. This check confirms both halves
of that statement.

```{r region-effect}
eu_placebo <- solve_arms(arm(REGION_EUROPE = 1))$pct_change_ldlc
eu_active  <- solve_arms(arm(REGION_EUROPE = 1, CONMED_ATORVASTATIN_DOSE = 10))$pct_change_ldlc
na_active  <- solve_arms(arm(CONMED_ATORVASTATIN_DOSE = 10))$pct_change_ldlc

cat(sprintf("Placebo arm: North America %+.3f%%, Europe %+.3f%% (shift %+.3f)\n",
            placebo, eu_placebo, eu_placebo - placebo))
cat(sprintf("Active-minus-placebo contrast: North America %+.4f%%, Europe %+.4f%%\n",
            na_active - placebo, eu_active - eu_placebo))

stopifnot(
  abs((eu_placebo - placebo) - (-4)) < 1e-8,                 # exactly -4 points
  abs((na_active - placebo) - (eu_active - eu_placebo)) < 1e-8  # cancels exactly
)
```

# Assumptions and deviations

* **No NCA / PKNCA validation.** This model has no PK layer, no concentrations
  and no time course; it maps a per-arm daily dose directly to a per-arm mean
  percentage change in LDL-C. Non-compartmental analysis is not defined for it.
  The validation above is instead a set of deterministic replications of the
  paper's own published outputs (Table 3 in full, the Discussion's point
  predictions and relative potencies, and the independence identity the paper
  states in words). This follows the same approach as the sibling MBMA
  vignettes `Vargo_2014_statins_ezetimibe_mbma` and
  `Mandema_2011_anticoagulants_mbma`.

* **Residual SD held at zero.** Equation 1 declares a between-subject residual
  `eps ~ N(0, sigma^2)`, but **no numeric value for sigma is published
  anywhere** in the paper -- not in Table 2, not in the Results text, and there
  is no supplement. Rather than invent a variance, `addSd` is wrapped in
  `fixed(0)`. The endpoint is still declared, so a downstream user can unfix and
  estimate it against their own data; as shipped, the model returns the typical
  study-arm mean exactly. Note that in the source the residual applied to a
  study-arm *mean*, so it would in any case need sample-size reweighting before
  being used as an individual-level error.

* **Trial random effect not encoded.** Equation 1 includes a trial-specific
  random effect `eta` with variance `nu^2`. The paper reports that this variance
  "was found to be not statistically significantly different from zero,
  suggesting homogeneity between studies," and gives no numeric estimate. It is
  therefore omitted rather than encoded as a zero-variance eta, which would
  produce a singular OMEGA that rxode2's Cholesky sampler cannot decompose.
  The sibling models `Vargo_2014_statins_ezetimibe_mbma` and
  `Mandema_2011_anticoagulants_mbma` treat their driven-to-zero between-trial
  variances the same way.

* **Uncertainty bands not reproduced.** The paper's 5th/95th percentiles in
  Table 3 and Figure 6 come from sampling 5,000 parameter sets from the
  estimated variance-covariance matrix. Table 2 publishes only marginal 95%
  confidence intervals, not the covariance matrix, so those bands cannot be
  reconstructed faithfully. Only the means are validated above. Treating the
  marginal intervals as independent would understate the correlation between
  Emax and ED50 and give misleadingly wide bands.

* **European placebo effect is prose-only, to one significant figure.** The
  value -4 percentage points comes from the Results sentence "An additional 4%
  reduction in LDL-C was consistently observed in Europeans trials." It has no
  row in Table 2 and no confidence interval. The reference category is taken to
  be North American (United States) trials, which is how the sentence reads
  ("a difference between North American and European studies ... an *additional*
  4% reduction ... in European trials") and is consistent with Table 2's
  unqualified `E0`. Because the paper does not tabulate which trials were
  European, this covariate cannot be cross-checked against any published output.

* **Only the LDL-C endpoint is extracted.** The Abstract and Introduction also
  describe models for HDL-C, persistent alanine aminotransferase elevation,
  myalgia, headache, and coronary-artery-disease risk reduction. The paper
  states that "the LDL-C effect ... is the main focus of this article" and
  publishes parameter estimates for the LDL-C dose-response only (Table 2). No
  parameter values for the other endpoints appear anywhere in the paper, so
  they cannot be extracted; this is a reporting gap in the source, not a scope
  decision.

* **Two nonstatins written as two terms.** Equation 1 has a single
  `Enonstatin` and a single `g` because the paper only ever combined a statin
  with *one* nonstatin. The model file writes out one term per nonstatin, each
  with its own interaction coefficient. For every arm the paper actually
  studied this reduces exactly to Equation 1. An arm coding both ezetimibe and
  gemcabene at once would add their effects with no ezetimibe-gemcabene
  interaction term -- a combination the paper never studied and did not
  parameterise.

* **Statin selection by summed potency ratio.** Equation 2 is
  `Dose^n * Emax / (Dose^n + ED50^n)`, which is identically
  `Emax * ratio^n / (ratio^n + 1)` for a potency-normalised dose
  `ratio = Dose / ED50`. Since the five statins share `Emax` and `n`, the model
  sums those potency-normalised doses across statins. Every trial arm in Table 1
  contains at most one statin, so exactly one term is non-zero and this is
  exact. Coding two statins simultaneously is outside the paper's calibration
  range.

* **Species and population.** Human. The pooled cohort is predominantly adults
  with hypercholesterolemia, but three of the four gemcabene trials enrolled
  healthy volunteers, healthy obese subjects, or subjects with low HDL-C; their
  much lower baseline LDL-C was adjusted for by an upstream ANOVA before the
  arm means entered this dose-response model, so that adjustment is baked into
  the source data rather than into the model equations.
