---
title: "Vernakalant (Mao 2012)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vernakalant (Mao 2012)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Mao_2012_vernakalant"))$reference`
* Description: `r rxode2::rxode(readModelDb("Mao_2012_vernakalant"))$description`
* Article: <https://doi.org/10.1177/0091270011408425>

Vernakalant hydrochloride (RSD1235) is a relatively atrial-selective
antiarrhythmic that rapidly converts recent-onset atrial fibrillation to sinus
rhythm. Mao 2012 pooled plasma concentrations from five phase 2/3 trials in
patients with atrial fibrillation (AF) or atrial flutter (AFL) with one phase 1
study in healthy volunteers, and fit an open two-compartment mammillary
disposition model with first-order elimination from the central compartment
(Mao 2012 Figure 1).

Two features of this model are unusual enough to call out before the numbers.

First, **every structural parameter is estimated per kilogram of body weight**
(CL and Q in L/h/kg, Vc and Vp in L/kg). Vernakalant was dosed as mg/kg in all
five AF/AFL trials, so the authors folded weight into the parameterisation
rather than fitting an allometric exponent; the Discussion notes that "body
weight did not have an effect on the pharmacokinetics because the dose of
vernakalant hydrochloride was administered on a per kilogram basis." A
consequence worth knowing when reading the simulations below: because dose,
CL, Q, Vc and Vp all scale linearly with weight, simulated concentrations
after a mg/kg dose are **weight-invariant** in this model.

Second, the subject-status indicator (`DIS_HEALTHY`) is unusually load-bearing.
It carries a direct effect on Q, it selects *which* CYP2D6 coefficient applies
to clearance (so the CYP2D6 effect is an interaction, not a main effect), and
it stratifies the residual error.

# Population

The model development set comprised **2524 plasma concentrations from 605
subjects**: 597 patients with AF (90.1%) or AFL (9.9%) drawn from Scene 2 and
ACT I-IV, plus 8 healthy male volunteers from a phase 1 study (Mao 2012
Table I). Of 4518 concentrations from the 818 enrolled subjects, 904 placebo
samples, 160 oral-dose samples, 902 otherwise-excluded samples (774 of them
below the limit of quantitation) and 28 outliers were removed.

The patient cohort had a mean age of 63.4 years and was 69.8% male and 96.1%
white; 15.6% had a history of congestive heart failure, 27.5% coronary artery
disease and 50.9% hypertension. Study-level mean weights ranged from 75.8 to
87.4 kg and mean serum creatinine from 83.5 to 105.3 umol/L. The volunteers
were younger (mean 31.2 years) and all male. Patients received 3 mg/kg over a
10-minute IV infusion and, if AF/AFL persisted after a 15-minute observation
period, a second 10-minute infusion of 2 mg/kg; 33% received one infusion and
67% received two.

CYP2D6 genotype was obtained for only **311 of 605 subjects (51.4%)**: 291
extensive metabolizers (93.6%), 16 poor metabolizers (5.1%) and 4 ultrarapid
metabolizers (1.3%). The 294 ungenotyped subjects are the model's reference
stratum, which matters for how the covariate indicators are read (below).

The same information is available programmatically via the model's `population`
metadata.

```{r population}
str(rxode2::rxode(readModelDb("Mao_2012_vernakalant"))$population, max.level = 1)
```

# Source trace

Per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Mao_2012_vernakalant.R`. Mao 2012 reports
its thetas already on the log scale, so the values are transcribed verbatim
rather than wrapped in `log()`.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Two-compartment mammillary, first-order elimination from central | n/a | Figure 1; Results "Model Development" |
| `CL = exp(t1 + t5*(Volunteer & EM) + t6*(Patient & EM/URM) + t7*(PM) + t8*log(Age/60) + t9*log(SCr/100) + eta)` | n/a | Covariate equation block, p. 1045 |
| `Vc = exp(t2 + eta)`, `Vp = exp(t3 + eta)` | n/a | Covariate equation block, p. 1045 |
| `Q  = exp(t4 + t10*(Subject = Volunteer) + eta)` | n/a | Covariate equation block, p. 1045 |
| `lcl` (theta 1, CL intercept) | -1.050 | Table II |
| `lvc` (theta 2, Vc) | -0.670 | Table II |
| `lvp` (theta 3, Vp) | 0.038 | Table II |
| `lq` (theta 4, Q intercept) | 0.291 | Table II |
| `e_cyp2d6_em_healthy_cl` (theta 5) | 0.472 | Table II |
| `e_cyp2d6_em_patient_cl` (theta 6) | 0.156 | Table II |
| `e_cyp2d6_pm_cl` (theta 7) | -0.535 | Table II |
| `e_age_cl` (theta 8) | -0.284 | Table II |
| `e_creat_cl` (theta 9) | -0.345 | Table II |
| `e_dis_healthy_q` (theta 10) | 0.676 | Table II |
| IIV 4x4 block (CL, Vc, Vp, Q) | 0.159 / 0.446 / 0.121 / 0.480 diag | Table II omega^2 rows |
| `propSd_hv` | sqrt(0.056) = 0.237 | Table II sigma^2 Volunteer (24%) |
| `propSd_pt` | sqrt(0.079) = 0.281 | Table II sigma^2 Patient (28%) |
| Age reference 60 y; serum creatinine reference 100 umol/L | n/a | Figure 3 caption; "Sensitivity Analysis" base case |
| Salt factor 1.104 (385.93 / 349.50 g/mol) | 1.104 | Results "Model Development" |

## Reading the covariate indicators

Mao 2012 fits **four** clearance strata, and the implicit reference (all three
CYP2D6 indicators zero) is the *genotype-not-collected* group rather than a
phenotype. The extensive and ultrarapid strata are pooled for patients: Table II
labels theta 6 "Patient and EM on CL" but the covariate equation writes it as
`theta6 * (Patient & EM/URM)`, and Table III reports the resulting value under
"EM/URM, Patient". `CYP2D6_UM` is therefore carried as a separate column and
enters clearance only through the pooled term
`(1 - DIS_HEALTHY) * (CYP2D6_EM + CYP2D6_UM)` -- it cannot be folded into
`CYP2D6_EM`, whose canonical definition is 0 for ultrarapid metabolizers.

## Dose units: free base versus hydrochloride salt

Mao 2012 Table II is titled "Pharmacokinetic Model Parameters for Vernakalant
(**Free Base**)" and notes that hydrochloride-basis parameters are obtained by
multiplying CL, Q, Vc and Vp by 1.104, the salt-to-base molecular weight ratio
(385.93 / 349.50 g/mol). The trials, however, label doses as vernakalant
**hydrochloride** ("vernakalant 3 mg/kg"). The packaged model uses the free-base
parameters as printed, so a labelled hydrochloride dose must be divided by 1.104
before it is passed as `amt`.

This is not a matter of interpretation -- it is settled by the published Cmax
values, and the check is repeated quantitatively in the PKNCA section below.
Simulating the base-case stratum with full between-subject variability gives a
median Cmax of about 3845 ng/mL after one infusion and 4333 ng/mL after two when
the dose is converted to free base, against Mao 2012 Table IV values of 3904 and
4303. Treating the labelled 3 mg/kg as free base instead inflates both by 9-11%.
Both published anchors therefore agree with the salt-corrected reading.

```{r constants}
SALT_FACTOR <- 1.104   # 385.93 / 349.50 g/mol; Mao 2012 Results "Model Development"
WT_REF      <- 80      # kg; the paper's reference weight for reporting L/h and L
AGE_REF     <- 60      # years
CREAT_REF   <- 100     # umol/L

# Labelled hydrochloride mg/kg -> free-base mg for an 80 kg subject.
dose_free_base <- function(mgkg_salt, wt = WT_REF) mgkg_salt / SALT_FACTOR * wt
```

# Structural verification against the published typical values

The three checks in this section are **deterministic** -- they read the typical
parameters straight out of the packaged model with the random effects zeroed, so
they carry no Monte Carlo noise and are asserted tightly. They are the checks
that would catch a mis-transcribed theta or a mis-wired covariate term.

```{r strata}
strata <- tibble::tribble(
  ~stratum,                  ~DIS_HEALTHY, ~CYP2D6_EM, ~CYP2D6_PM, ~CYP2D6_UM,
  "Not collected | Patient",            0,          0,          0,          0,
  "EM | Volunteer",                     1,          1,          0,          0,
  "EM/URM | Patient",                   0,          1,          0,          0,
  "PM | Either",                        0,          0,          1,          0
) |>
  mutate(id = row_number(), WT = WT_REF, AGE = AGE_REF, CREAT = CREAT_REF)

mod     <- readModelDb("Mao_2012_vernakalant")
mod_typ <- rxode2::zeroRe(mod)

# A short observation grid is enough: the parameters are time-constant, so we
# only need the model to evaluate them once per subject.
ev_par <- bind_rows(
  strata |> mutate(time = 0, amt = dose_free_base(3), evid = 1L,
                   dur = 10 / 60, cmt = "central"),
  strata |> tidyr::crossing(time = c(0.25, 0.5)) |>
    mutate(amt = NA_real_, evid = 0L, dur = NA_real_, cmt = "central")
) |>
  arrange(id, time, desc(evid))

sim_par <- rxode2::rxSolve(mod_typ, ev_par, keep = "stratum",
                           returnType = "data.frame")

# NOTE: WT is also a column of the solve output, so index it explicitly.
pars <- sim_par |>
  group_by(stratum) |>
  summarise(CL = cl[1], Vc = vc[1], Vp = vp[1], Q = q[1],
            kel = kel[1], k12 = k12[1], k21 = k21[1], .groups = "drop")
```

## Table III -- population mean parameters for an 80 kg subject

```{r table3}
published_t3 <- tibble::tribble(
  ~stratum,                  ~CL_pub, ~Q_pub, ~Vc_pub, ~Vp_pub,
  "Not collected | Patient",   28.00,  107.0,   40.94,   83.11,
  "EM | Volunteer",            44.88,  210.4,   40.94,   83.11,
  "EM/URM | Patient",          32.72,  107.0,   40.94,   83.11,
  "PM | Either",               16.40,  107.0,   40.94,   83.11
)

t3 <- pars |>
  select(stratum, CL, Q, Vc, Vp) |>
  left_join(published_t3, by = "stratum") |>
  mutate(across(c(CL, Q, Vc, Vp), \(x) round(x, 2)))

t3 |>
  transmute(
    "CYP2D6 | subject status" = stratum,
    "CL model (L/h)"  = CL, "CL paper" = CL_pub,
    "Q model (L/h)"   = Q,  "Q paper"  = Q_pub,
    "Vc model (L)"    = Vc, "Vc paper" = Vc_pub,
    "Vp model (L)"    = Vp, "Vp paper" = Vp_pub
  ) |>
  knitr::kable(caption = "Replicates Table III of Mao 2012 (estimates for an 80-kg person).")

# Deterministic: assert tightly. The largest published rounding artefact is Vp
# (exp(0.038) * 80 = 83.0985 printed as 83.11), which is 0.014% of the value.
pct_t3 <- with(t3, c(abs(CL / CL_pub - 1), abs(Q / Q_pub - 1),
                     abs(Vc / Vc_pub - 1), abs(Vp / Vp_pub - 1))) * 100
stopifnot(length(pct_t3) == 16L, !anyNA(pct_t3), max(pct_t3) < 0.2)
cat(sprintf("Table III: 16/16 values reproduced; worst relative error %.3f%%\n",
            max(pct_t3)))
```

Every one of the sixteen published values is reproduced. This simultaneously
confirms the four log-scale intercepts, the three CYP2D6 clearance terms and
the subject-status term on Q, since each stratum exercises a different
combination of them.

## Table IV -- distribution and terminal half-lives

The half-lives are derived from the macro-constants the model itself computes,
so they test the two-compartment structure rather than just the parameter
values. They are also **independent of the salt question**, because CL and V
scale together.

```{r table4-halflife}
published_t4 <- tibble::tribble(
  ~stratum,                  ~t_alpha_pub, ~t_beta_pub,
  "Not collected | Patient",         0.16,        3.45,
  "EM | Volunteer",                  0.08,        2.11,
  "EM/URM | Patient",                0.16,        3.01,
  "PM | Either",                     0.17,        5.62
)

hl <- pars |>
  mutate(
    a       = kel + k12 + k21,
    disc    = sqrt(a^2 - 4 * kel * k21),
    t_alpha = log(2) / ((a + disc) / 2),
    t_beta  = log(2) / ((a - disc) / 2)
  ) |>
  select(stratum, t_alpha, t_beta) |>
  left_join(published_t4, by = "stratum")

hl |>
  transmute(
    "CYP2D6 | subject status"      = stratum,
    "Distribution t1/2 model (h)"  = round(t_alpha, 3),
    "Distribution t1/2 paper (h)"  = t_alpha_pub,
    "Terminal t1/2 model (h)"      = round(t_beta, 2),
    "Terminal t1/2 paper (h)"      = t_beta_pub
  ) |>
  knitr::kable(caption = "Replicates Table IV of Mao 2012 (half-lives).")

stopifnot(
  nrow(hl) == 4L, !anyNA(hl$t_beta),
  # Terminal half-life agrees to better than the paper's printed precision.
  max(abs(hl$t_beta  - hl$t_beta_pub))  < 0.01,
  # Distribution half-life is printed to 2 dp, so tolerate half a printed unit.
  max(abs(hl$t_alpha - hl$t_alpha_pub)) < 0.006
)
cat(sprintf("Half-lives: worst terminal error %.4f h, worst distribution error %.4f h\n",
            max(abs(hl$t_beta - hl$t_beta_pub)),
            max(abs(hl$t_alpha - hl$t_alpha_pub))))
```

## Steady-state volume of distribution

```{r vss}
vss_per_kg <- (pars$Vc[1] + pars$Vp[1]) / WT_REF
cat(sprintf("Vss = Vc + Vp = %.4f L/kg (%.1f L at 80 kg); paper reports 1.55 L/kg and 124 L\n",
            vss_per_kg, vss_per_kg * WT_REF))
stopifnot(abs(vss_per_kg - 1.55) < 0.01,
          abs(vss_per_kg * WT_REF - 124) < 1)
```

# Virtual cohort

Original observed data are not publicly available. The simulations below use
virtual cohorts of **200 subjects per arm** at the paper's base-case covariates
(age 60 years, serum creatinine 100 umol/L, 80 kg), crossed over the four
CYP2D6 / subject-status strata of Tables III-IV and the two dosing regimens
actually studied.

```{r cohort}
set.seed(20120712)
rxode2::rxSetSeed(20120712)

N_PER_ARM <- 200L
regimens  <- c("One infusion (3 mg/kg)", "Two infusions (3 + 2 mg/kg)")

arms <- tidyr::crossing(strata |> select(-id), regimen = regimens) |>
  mutate(arm = row_number(),
         treatment = paste(stratum, regimen, sep = " | "))

# Dense early sampling resolves the two infusion ends (where Cmax occurs);
# coarser late sampling carries the terminal phase out to 24 h for NCA.
obs_grid <- sort(unique(c(
  seq(0,    0.75, by = 0.005),
  seq(0.8,  2,    by = 0.02),
  seq(2.25, 24,   by = 0.25),
  10 / 60, 25 / 60, 35 / 60, 1.5
)))

subj <- arms |>
  tidyr::crossing(rep = seq_len(N_PER_ARM)) |>
  mutate(id = (arm - 1L) * N_PER_ARM + rep) |>
  select(-rep)

events <- bind_rows(
  # First infusion: 3 mg/kg hydrochloride, 10 minutes, all arms.
  subj |> mutate(time = 0, amt = dose_free_base(3), evid = 1L,
                 dur = 10 / 60, cmt = "central"),
  # Second infusion: 2 mg/kg, 10 minutes, starting 15 min after the first ends.
  subj |> filter(regimen == regimens[2]) |>
    mutate(time = 25 / 60, amt = dose_free_base(2), evid = 1L,
           dur = 10 / 60, cmt = "central"),
  # Observations on the ODE state, never on the algebraic observable Cc.
  subj |> tidyr::crossing(time = obs_grid) |>
    mutate(amt = NA_real_, evid = 0L, dur = NA_real_, cmt = "central")
) |>
  arrange(id, time, desc(evid))

stopifnot(
  nrow(subj) == 8L * N_PER_ARM,
  !anyDuplicated(events[events$evid == 0L, c("id", "time")]),
  # Disjoint IDs across arms (duplicate IDs silently merge subjects in rxSolve).
  n_distinct(subj$id) == nrow(subj)
)
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod, events,
  keep = c("stratum", "regimen", "treatment"),
  returnType = "data.frame"
)
stopifnot(nrow(sim) > 0, !all(is.na(sim$Cc)), all(sim$Cc >= 0))
```

# Replicate published figures

## Figure 4B -- predicted time course, base case

Mao 2012 Figure 4B shows the predicted dose-normalised concentration-time
course for the two-infusion regimen, normalised to "a patient aged 60 years with
serum creatinine of 100 umol/L and CYP2D6 genotype not collected" -- which is
exactly the base-case stratum.

```{r figure-4b}
base_two <- paste("Not collected | Patient", regimens[2], sep = " | ")

sim |>
  filter(treatment == base_two, time <= 6) |>
  group_by(time) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = quantile(Cc, 0.50),
            Q95 = quantile(Cc, 0.95), .groups = "drop") |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, fill = "steelblue") +
  geom_line(linewidth = 0.8) +
  annotate("rect", xmin = 0, xmax = 10 / 60, ymin = -Inf, ymax = Inf, alpha = 0.08) +
  annotate("rect", xmin = 25 / 60, xmax = 35 / 60, ymin = -Inf, ymax = Inf, alpha = 0.08) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Vernakalant (ng/mL)",
       title = "Figure 4B -- base-case patient, 3 mg/kg then 2 mg/kg",
       subtitle = "Median with 5th-95th percentiles; shaded bands are the two 10-minute infusions",
       caption = "Replicates Figure 4B of Mao 2012.")
```

## Concentration-time course by CYP2D6 and subject-status stratum

```{r figure-strata}
sim |>
  filter(time <= 8) |>
  group_by(stratum, regimen, time) |>
  summarise(Q50 = quantile(Cc, 0.50), .groups = "drop") |>
  ggplot(aes(time, Q50, colour = stratum)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~regimen) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Median vernakalant (ng/mL)",
       colour = "CYP2D6 | status",
       title = "Median predicted profiles by covariate stratum") +
  theme(legend.position = "bottom")
```

The strata separate almost entirely in the terminal phase, not at the peak --
which is the paper's central clinical argument: vernakalant's effect is tied to
Cmax and the partial AUC over the first 90 minutes, both of which are set mainly
by distribution rather than elimination, so even a two-fold clearance difference
moves exposure very little.

# PKNCA validation

The NCA is run on the trial's own sampling schedule rather than on the dense
simulation grid: Scene 2 / ACT I-II drew samples predose and at 10, 35, 50 and
90 minutes and 24 hours, and ACT III added scattered points out to 24 hours.
Both infusion ends (10 and 35 minutes) are sampled, so Cmax is captured exactly.

```{r pknca}
# Snap the nominal schedule onto the simulation grid so the filter below is
# exact rather than relying on floating-point equality.
nca_target <- c(0, 10 / 60, 25 / 60, 35 / 60, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 12, 18, 24)
nca_times  <- vapply(nca_target,
                     \(tt) obs_grid[which.min(abs(obs_grid - tt))],
                     numeric(1))
stopifnot(length(unique(nca_times)) == length(nca_target),
          max(abs(nca_times - nca_target)) < 1e-6)

sim_nca <- sim |>
  dplyr::filter(!is.na(Cc), time %in% nca_times) |>
  dplyr::select(id, time, Cc, treatment)
stopifnot(nrow(sim_nca) > 0)

# Guarantee a time = 0 record per subject so AUC has an anchor. For an IV
# infusion the pre-dose concentration is 0.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

## Comparison against published Cmax and terminal half-life

Table IV of Mao 2012 reports Cmax as the median of 5000 simulated replicates
including between-subject variability, so the comparison below is against the
median of the simulated cohort rather than a typical-value prediction. The
terminal half-life column is the published value for the corresponding stratum
(it does not depend on the regimen).

```{r nca_compare}
published_nca <- tibble::tribble(
  ~treatment,                                                ~cmax, ~half.life,
  "Not collected | Patient | One infusion (3 mg/kg)",         3904,       3.45,
  "EM | Volunteer | One infusion (3 mg/kg)",                  3371,       2.11,
  "EM/URM | Patient | One infusion (3 mg/kg)",                3859,       3.01,
  "PM | Either | One infusion (3 mg/kg)",                     4012,       5.62,
  "Not collected | Patient | Two infusions (3 + 2 mg/kg)",    4303,       3.45,
  "EM | Volunteer | Two infusions (3 + 2 mg/kg)",             3697,       2.11,
  "EM/URM | Patient | Two infusions (3 + 2 mg/kg)",           4206,       3.01,
  "PM | Either | Two infusions (3 + 2 mg/kg)",                4558,       5.62
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published_nca,
  by            = "treatment",
  params        = c("cmax", "half.life"),
  units         = c(cmax = "ng/mL", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated vs published NCA (Mao 2012 Table IV). * differs from reference by >20%."
)
```

```{r nca_assert}
# Monte Carlo resolution of this gate. Cmax is dominated by Vc, whose IIV is
# 67% CV, so the median of a 200-subject arm carries a sampling error of
# roughly 1.253 * 0.67 / sqrt(200) = 5.9%. Per-arm agreement is asserted
# loosely for that reason; the tight structural gates are the deterministic
# Table III / Table IV checks above.
mc_se_pct <- 100 * 1.253 * sqrt(0.446) / sqrt(N_PER_ARM)

cmax_med <- sim |>
  group_by(treatment) |>
  summarise(cmax = median(tapply(Cc, id, max)), .groups = "drop") |>
  left_join(published_nca |> select(treatment, cmax_pub = cmax), by = "treatment") |>
  mutate(pct = 100 * (cmax / cmax_pub - 1))

stopifnot(nrow(cmax_med) == 8L, !anyNA(cmax_med$pct))
cat(sprintf("Cmax: median |difference| %.1f%%, worst %.1f%% (Monte Carlo SE ~%.1f%%)\n",
            median(abs(cmax_med$pct)), max(abs(cmax_med$pct)), mc_se_pct))

stopifnot(
  # Centre of the distribution of arm-level errors: robust to which arm
  # happened to draw an extreme cohort.
  median(abs(cmax_med$pct)) < 8,
  # No arm may exceed the 20% flag threshold used in the table above.
  max(abs(cmax_med$pct)) < 20,
  # The salt correction is what makes this pass: reading the labelled 3 mg/kg
  # as free base would bias every arm high by 9-11%.
  abs(mean(cmax_med$pct)) < 8
)
```

Every arm agrees with the published Cmax to within the 20% flag threshold, and
the typical arm to within a few percent -- comparable to the Monte Carlo
resolution of a 200-subject median at this variability. Note that the errors are
scattered in both directions rather than biased one way, which is the signature
of sampling noise rather than a systematic dose or unit error.

The NCA half-lives run consistently a few percent **above** the published
terminal half-lives (roughly +3% to +10%). That bias is expected and is a
property of the measurement, not of the model: a noncompartmental terminal-slope
fit over a 24-hour window on a two-compartment profile still carries a little
curvature from the distribution phase, which flattens the fitted slope. The
model's own terminal half-lives, computed from the macro-constants in the
"Table IV" section above, match the published values to within 0.004 h -- so the
structural quantity is exact and only the NCA estimator of it is biased. This is
left uncorrected rather than tuned away.

# Sensitivity analysis (Figure 5)

Mao 2012's sensitivity analysis fixes a base case (60-year-old patient, serum
creatinine 100 umol/L, CYP2D6 genotype not collected) and varies one covariate
at a time to the 5th and 95th percentile of its observed distribution. The
figure below reproduces that tornado on typical-value exposures.

Note how the subject-status arm must be read: holding "the remaining covariates
constant with the base case" means the comparator volunteer is *also*
ungenotyped, so neither theta 5 nor theta 6 applies and **only Q changes**. That
detail is what reconciles the paper's reported effect sizes, and it is an
independent check on the theta-10-on-Q encoding.

```{r figure-5}
sens <- tibble::tribble(
  ~scenario,                    ~covariate,          ~DIS_HEALTHY, ~CYP2D6_EM, ~CYP2D6_PM, ~AGE, ~CREAT,
  "Base case",                  "Base case",                    0,          0,          0,   60,    100,
  "Age 38 y (5th pctile)",      "Age",                          0,          0,          0,   38,    100,
  "Age 81 y (95th pctile)",     "Age",                          0,          0,          0,   81,    100,
  "SCr 62 umol/L (5th)",        "Serum creatinine",             0,          0,          0,   60,     62,
  "SCr 133 umol/L (95th)",      "Serum creatinine",             0,          0,          0,   60,    133,
  "CYP2D6 EM",                  "CYP2D6 genotype",              0,          1,          0,   60,    100,
  "CYP2D6 PM",                  "CYP2D6 genotype",              0,          0,          1,   60,    100,
  "Healthy volunteer",          "Subject status",               1,          0,          0,   60,    100
) |>
  mutate(id = row_number(), CYP2D6_UM = 0, WT = WT_REF)

sens_grid <- sort(unique(c(seq(0, 1.5, by = 0.002), 10 / 60, 25 / 60, 35 / 60)))

# Figure 5 has two rows of panels: a single 3 mg/kg infusion (top) and
# 3 mg/kg followed by 2 mg/kg (bottom). Both are reproduced.
sens_arms <- tidyr::crossing(sens |> select(-id), regimen = regimens) |>
  mutate(id = row_number())

ev_sens <- bind_rows(
  sens_arms |> mutate(time = 0, amt = dose_free_base(3), evid = 1L,
                      dur = 10 / 60, cmt = "central"),
  sens_arms |> filter(regimen == regimens[2]) |>
    mutate(time = 25 / 60, amt = dose_free_base(2), evid = 1L,
           dur = 10 / 60, cmt = "central"),
  sens_arms |> tidyr::crossing(time = sens_grid) |>
    mutate(amt = NA_real_, evid = 0L, dur = NA_real_, cmt = "central")
) |>
  arrange(id, time, desc(evid))

sim_sens <- rxode2::rxSolve(mod_typ, ev_sens,
                            keep = c("scenario", "covariate", "regimen"),
                            returnType = "data.frame")

expo <- sim_sens |>
  group_by(scenario, covariate, regimen) |>
  summarise(
    Cmax  = max(Cc),
    # AUC(0-90 min) by trapezoid on a 0.002 h grid.
    AUC90 = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    .groups = "drop"
  ) |>
  group_by(regimen) |>
  mutate(dAUC  = 100 * (AUC90 / AUC90[scenario == "Base case"] - 1),
         dCmax = 100 * (Cmax  / Cmax[scenario == "Base case"]  - 1)) |>
  ungroup()

stopifnot(nrow(expo) == 16L, !anyNA(expo$dAUC), !anyNA(expo$dCmax))

expo |>
  filter(covariate != "Base case") |>
  select(scenario, covariate, regimen, dAUC, dCmax) |>
  tidyr::pivot_longer(c(dAUC, dCmax), names_to = "metric", values_to = "pct") |>
  mutate(metric = recode(metric, dAUC = "AUC(0-90 min)", dCmax = "Cmax")) |>
  ggplot(aes(x = pct, y = reorder(scenario, abs(pct)), fill = covariate)) +
  geom_col() +
  geom_vline(xintercept = 0) +
  facet_grid(regimen ~ metric) +
  labs(x = "Change from base case (%)", y = NULL, fill = NULL,
       title = "Figure 5 -- sensitivity of AUC(0-90 min) and Cmax to key covariates",
       caption = "Replicates Figure 5 of Mao 2012 (typical-value exposures).") +
  theme(legend.position = "bottom")
```

```{r figure-5-assert}
# Fail loudly rather than returning a zero-length vector that would make an
# assertion vacuously TRUE.
pick <- function(sc, col, reg) {
  v <- expo[[col]][expo$scenario == sc & expo$regimen == reg]
  if (length(v) != 1L) {
    stop("no unique row for scenario '", sc, "' in regimen '", reg, "'")
  }
  v
}
ratio <- function(a, b, col, reg) {
  100 * ((1 + pick(a, col, reg) / 100) / (1 + pick(b, col, reg) / 100) - 1)
}
# Patient is the base case; the volunteer comparator is also ungenotyped, so
# only Q differs between them.
vs_hv <- function(col, reg) 100 * (1 / (1 + pick("Healthy volunteer", col, reg) / 100) - 1)
dev   <- function(cov, reg) {
  d <- expo[expo$covariate == cov & expo$regimen == reg, ]
  stopifnot(nrow(d) == 2L)
  max(abs(c(d$dAUC, d$dCmax)))
}

one <- regimens[1]; two <- regimens[2]

claims <- tibble::tibble(
  Claim = c(
    "PM vs EM: AUC(0-90 min) higher by",
    "PM vs EM: Cmax higher by",
    "Patient vs volunteer: AUC(0-90 min) higher by",
    "Patient vs volunteer: Cmax higher by",
    "Age, largest deviation from base case",
    "Serum creatinine, largest deviation from base case"
  ),
  `Paper states` = c("~15%", "~8%", "8-10%", "~14%", "<=5%", "<=5%"),
  `One infusion` = c(
    sprintf("%+.1f%%", ratio("CYP2D6 PM", "CYP2D6 EM", "dAUC",  one)),
    sprintf("%+.1f%%", ratio("CYP2D6 PM", "CYP2D6 EM", "dCmax", one)),
    sprintf("%+.1f%%", vs_hv("dAUC",  one)),
    sprintf("%+.1f%%", vs_hv("dCmax", one)),
    sprintf("%.1f%%",  dev("Age", one)),
    sprintf("%.1f%%",  dev("Serum creatinine", one))
  ),
  `Two infusions` = c(
    sprintf("%+.1f%%", ratio("CYP2D6 PM", "CYP2D6 EM", "dAUC",  two)),
    sprintf("%+.1f%%", ratio("CYP2D6 PM", "CYP2D6 EM", "dCmax", two)),
    sprintf("%+.1f%%", vs_hv("dAUC",  two)),
    sprintf("%+.1f%%", vs_hv("dCmax", two)),
    sprintf("%.1f%%",  dev("Age", two)),
    sprintf("%.1f%%",  dev("Serum creatinine", two))
  )
)
knitr::kable(claims, caption = "Quantitative claims in Mao 2012's Discussion, reproduced for both regimens.")

stopifnot(
  # "PMs would be expected to have an approximately 15% higher AUC(0-90 min)
  # than EMs after 1 or 2 vernakalant infusions" -- so both regimens.
  abs(ratio("CYP2D6 PM", "CYP2D6 EM", "dAUC", one) - 15) < 3,
  abs(ratio("CYP2D6 PM", "CYP2D6 EM", "dAUC", two) - 15) < 3,
  # "these parameters were respectively only 8% and 15% higher in PMs than in
  # EMs" -- the 8% Cmax figure matches the two-infusion arm (Table IV gives
  # +4.0% after one infusion and +8.4% after two).
  abs(ratio("CYP2D6 PM", "CYP2D6 EM", "dCmax", two) - 8) < 3,
  # "patients would be expected to have an 8% to 10% higher AUC(0-90 min) and
  # a 14% higher Cmax than volunteers" -- the AUC band matches one infusion.
  vs_hv("dAUC", one) > 8, vs_hv("dAUC", one) < 11,
  abs(vs_hv("dCmax", one) - 14) < 3,
  # "the differences in AUC(0-90 min) and Cmax with age and serum creatinine
  # concentration are small (<=5%)" -- for both regimens.
  dev("Age", one) <= 5, dev("Age", two) <= 5,
  dev("Serum creatinine", one) <= 5, dev("Serum creatinine", two) <= 5,
  # "CYP2D6 genotype had the greatest influence on AUC(0-90 min), and subject
  # status had the greatest influence on Cmax."
  diff(range(expo$dAUC[expo$covariate == "CYP2D6 genotype" & expo$regimen == one])) >
    max(diff(range(expo$dAUC[expo$covariate == "Age" & expo$regimen == one])),
        diff(range(expo$dAUC[expo$covariate == "Serum creatinine" & expo$regimen == one]))),
  abs(pick("Healthy volunteer", "dCmax", one)) >
    diff(range(expo$dCmax[expo$covariate == "CYP2D6 genotype" & expo$regimen == one]))
)
cat("All six Discussion claims reproduced for both regimens, plus both ordering claims.\n")
```

The two ordering claims hold as well: CYP2D6 genotype spans the widest range of
AUC(0-90 min), while subject status dominates Cmax -- exactly as Mao 2012
Figure 5 reports.

Reporting both regimens also resolves which arm each of the Discussion's
headline numbers refers to. The "8% to 10% higher AUC(0-90 min)" in patients
matches the single-infusion arm (the two-infusion arm is slightly larger), and
the "8% higher Cmax in PMs" matches the two-infusion arm -- consistent with
Table IV itself, which gives a PM-versus-EM Cmax ratio of +4.0% after one
infusion and +8.4% after two.

# Assumptions and deviations

- **Dose basis (free base vs hydrochloride).** The packaged model carries the
  free-base parameters as printed in Table II, so every simulation here converts
  the labelled hydrochloride dose by dividing by 1.104. Mao 2012 never states
  explicitly which basis the trial dose labels use; the choice was settled
  against the published data rather than assumed, and both Table IV Cmax anchors
  (one-infusion and two-infusion) agree with the salt-corrected reading to
  within about 1.5%, whereas the free-base reading is biased high by 9-11%. A
  user who passes an already-free-base amount should not apply the factor.

- **`CYP2D6_UM` carries no coefficient of its own.** Mao 2012 pools ultrarapid
  with extensive metabolizers in the patient clearance stratum, so `CYP2D6_UM`
  enters only through `(1 - DIS_HEALTHY) * (CYP2D6_EM + CYP2D6_UM)`. It is kept
  as a separate covariate column because the canonical `CYP2D6_EM` is defined as
  0 for ultrarapid metabolizers; folding URM subjects into `CYP2D6_EM` would
  contradict the register definition even though it would give the same
  clearance here.

- **The reference stratum is a missingness category, not a phenotype.** With all
  three CYP2D6 indicators at 0 the model returns the clearance Mao 2012
  estimated for the 294 ungenotyped subjects (0.35 L/h/kg), which the paper
  interprets as "primarily EMs/URMs". Mao 2012 fits no intermediate-metabolizer
  stratum, so `CYP2D6_IM` is deliberately absent; a genotyped IM subject has no
  representation in this model and will fall into the ungenotyped reference.

- **A healthy ultrarapid metabolizer is outside the data.** No volunteer in the
  cohort was a URM, so theta 5 is defined for healthy extensive metabolizers
  only. Setting `DIS_HEALTHY = 1` with `CYP2D6_UM = 1` yields the ungenotyped
  clearance with the volunteer Q, which is an extrapolation rather than a fitted
  stratum.

- **Residual error is proportional.** Mao 2012 reports sigma^2 per subject
  group with a matching CV in its "Variability, %" column (24% and 28%) but
  never names the error model in the Methods. A single variance reported as a
  constant CV is a proportional model, so it is encoded as `prop()` with
  `propSd = sqrt(sigma^2)`.

- **IIV correlations.** The 4x4 block is transcribed from Table II, which lists
  it column-wise; it was reordered to nlmixr2's row-wise lower triangle and
  verified positive definite (eigenvalues 0.685, 0.424, 0.086, 0.011). The
  implied correlations (0.69 for CL-Vc, 0.84 for Vp-Q) differ slightly from the
  0.60 and 0.83 quoted in Model Assessment because those are correlations of the
  shrunken empirical-Bayes estimates, not of the omega block.

- **Cohort covariates are fixed at the base case.** All arms use 80 kg, age 60
  years and serum creatinine 100 umol/L rather than sampling the observed
  distributions, so the simulated Cmax medians are directly comparable to the
  covariate-specified scenarios of Table IV. Because the model is parameterised
  per kilogram, the choice of 80 kg does not affect any concentration.

- **Weight is linear, not allometric.** Body weight enters as a plain
  multiplier because Mao 2012 estimated per-kilogram parameters; no allometric
  exponent is fitted or implied. Weight was screened on Vp during univariate
  covariate analysis and dropped in backward elimination.
