---
title: "Ceftriaxone (Marianski 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ceftriaxone (Marianski 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Marianski_2026_ceftriaxone"))$reference`
* Description: `r rxode2::rxode(readModelDb("Marianski_2026_ceftriaxone"))$description`
* Article: <https://doi.org/10.1093/ofid/ofaf695.1445>
* Supplement: none. Europe PMC reports `hasSuppl = N` for PMC12792806 and the
  abstract cites no supplementary material.

Marianski 2026 is an IDWeek 2025 **poster abstract** (P-1254, Session 148,
PK/PD Studies) reporting a two-compartment population PK model for intravenous
ceftriaxone in 44 critically ill children with multiple organ dysfunction
syndrome, together with a free-time-over-MIC (fT>MIC) target-attainment
analysis. It is a conference abstract rather than a peer-reviewed paper, but it
is completely encodable: Table 1 publishes all four structural estimates, all
four interindividual variances, the residual-error coefficient and the four
covariate equations.

## The parameter table is a raster image

Every number this model needs lives inside Table 1, and Table 1 is published as
an **embedded JPEG**. `pdftotext` recovers the abstract prose and drops the
entire table, so a text-only read of this paper makes it look like an
unencodable reporting-gap abstract. Recovering the panels is one command:

```
pdfimages -png -f 1 -l 1 unknown_2026_PMC12792806.pdf pg
```

which yields `pg-000.png` (Table 1), `pg-001.png` (Figure 1, goodness of fit)
and `pg-002.png` (Figure 2, fT>MIC by MIC). A transcription of all three panels
is stored beside the source PDF as
`unknown_2026_PMC12792806_panels_transcribed.md`.

The PDF page carries **two unrelated abstracts in two columns**, so panel
ownership was proved rather than assumed, on three independent matches:

1. Table 1's random-effects `C.V.(%)` values of 43.11 and 45.45 reproduce the
   Results sentence "Between subject variation (CV%) for V1 and CL was 43.1%
   and 45.5%, respectively (Table 1)" exactly.
2. Figure 2's caption names "CRO MIC" and the 90% protein binding that this
   abstract's Methods introduce.
3. Table 1's caption names the 27.4 kg weight normalisation used in its own
   equations.

The neighbouring abstract's prose (hyperfiltration, BMI, sex and "nature of the
hematologic disease" not correlated) belongs to a **different** study and none of
it is attributed to this model.

# Population

Forty-four patients aged 2 months to 17 years were enrolled in a multi-centre
prospective study of antibiotic PK (grant R01HD103755, PALISI network) in
critically ill children under 18 years with multiple organ dysfunction
syndrome, defined as two or more organ failures, who were prescribed
ceftriaxone as standard of care. Children on extracorporeal support were
**excluded**, so the model carries no information about ECMO or renal
replacement therapy.

Up to 15 PK samples per patient were collected over 3 days by volumetric
absorptive microsampling (VAMS, 20 uL per sample) and quantified in **whole
blood** by a validated LC-MS/MS assay. Estimated GFR by the CKiD "under 25"
(U25) equation ranged from 25 to 225 mL/min/1.73 m^2 with a median of 100,
spanning moderate renal impairment through frank hyperfiltration.

```{r population}
pop <- rxode2::rxode(readModelDb("Marianski_2026_ceftriaxone"))$population
tibble::tibble(Field = names(pop), Value = vapply(pop, function(x)
  paste(as.character(x), collapse = "; "), character(1))) |>
  knitr::kable(caption = "Population metadata recorded in the model file.")
```

Two demographic facts the abstract **does not** report are load-bearing for any
simulation and are assumed here (see Assumptions and deviations): the cohort
**weight distribution**, and the **dosing regimen**.

# Source trace

```{r source-trace}
tibble::tribble(
  ~Quantity, ~Value, ~`Source location`,
  "V1 (central volume)",            "8.22 L",   "Table 1, Fixed Effects row 'V1 (L)'",
  "Q (intercompartmental)",         "0.75 L/h", "Table 1, Fixed Effects row 'Q (L/hr)'",
  "V2 (peripheral volume)",         "13.74 L",  "Table 1, Fixed Effects row 'V2 (L)'",
  "CL (clearance)",                 "0.91 L/h", "Table 1, Fixed Effects row 'CL (L/hr)'",
  "IIV V1",                         "SD 0.41, CV 43.11%",  "Table 1, Random Effects row 'V1'",
  "IIV Q",                          "SD 0.93, CV 117.84%", "Table 1, Random Effects row 'Q'",
  "IIV V2",                         "SD 1.29, CV 205.6%",  "Table 1, Random Effects row 'V2'",
  "IIV CL",                         "SD 0.43, CV 45.45%",  "Table 1, Random Effects row 'CL'",
  "Proportional residual error",    "b = 0.2",  "Table 1, Error Model Parameters row 'b'",
  "CL covariate equation",          "CL_pop * (WT/27.4)^0.75 * (GFR/120)", "Table 1, 'Model parameterized as:' block",
  "V1 covariate equation",          "V1_pop * (WT/27.4)^1", "Table 1, 'Model parameterized as:' block",
  "Q covariate equation",           "Q_pop (no covariate)",  "Table 1, 'Model parameterized as:' block",
  "V2 covariate equation",          "V2_pop (no covariate)", "Table 1, 'Model parameterized as:' block",
  "Free fraction for fT>MIC",       "0.10 (90% protein binding)", "Methods; Figure 2 caption",
  "eGFR equation",                  "CKiD U25",  "Methods",
  "eGFR range / median",            "25-225 / 100 mL/min/1.73 m^2", "Results",
  "Age range",                      "2 months to 17 years", "Results",
  "Estimation software",            "Monolix 2024R1", "Methods"
) |>
  knitr::kable(caption = "Source location for every model equation and ini() parameter.")
```

# Gate 1: the paper published its own answer key

Table 1 prints `Value`, `S.E.` and `R.S.E.(%)` for all nine estimated
quantities. Because `R.S.E.(%) = 100 * S.E. / Value` by definition, the table
over-determines itself: a mis-read digit in any of the 27 numbers breaks the
identity. This is a direct check that the raster was transcribed correctly.

```{r gate-transcription}
tab1 <- tibble::tribble(
  ~Row,          ~Value, ~SE,     ~RSE,
  "V1",          8.22,   0.74,    9.01,
  "Q",           0.75,   0.2,     26.3,
  "V2",          13.74,  4.56,    33.2,
  "CL",          0.91,   0.083,   9.18,
  "omega V1",    0.41,   0.064,   15.6,
  "omega Q",     0.93,   0.23,    24.3,
  "omega V2",    1.29,   0.25,    19.3,
  "omega CL",    0.43,   0.062,   14.3,
  "b (prop err)", 0.2,   0.0091,  4.54
) |>
  mutate(`RSE recomputed` = round(100 * SE / Value, 2),
         `Deviation (pp)` = round(abs(`RSE recomputed` - RSE), 3))

knitr::kable(tab1, caption = "R.S.E. recomputed from S.E. / Value for every Table 1 row.")

# The two largest deviations are the omega-Q and Q rows, whose S.E. is printed
# to only one or two significant figures (0.2 and 0.23); the rounding envelope
# fully accounts for them. 0.5 pp leaves headroom over that rounding while
# still failing on any single mis-read digit (which moves RSE by >= 1 pp).
stopifnot(max(tab1$`Deviation (pp)`) < 0.5)
```

# Gate 2: the omega convention is pinned by the table itself

The single most common way to mis-encode a published popPK model is to read an
interindividual-variability column on the wrong scale. Table 1 removes the
ambiguity by reporting each random effect **twice** -- as `SD` and as
`C.V.(%)`. The SD is the log-scale omega; the CV is the log-normal CV derived
from it. `nlmixr2` stores variances, so the model file carries
`omega^2 = log(CV^2 + 1)`.

```{r gate-omega}
printed <- tibble::tribble(
  ~Row,       ~`SD printed`, ~`CV printed`, ~`omega2 in ini()`,
  "omega V1", 0.41,  43.11,  0.170457,
  "omega Q",  0.93,  117.84, 0.870719,
  "omega V2", 1.29,  205.60, 1.653864,
  "omega CL", 0.43,  45.45,  0.187782
) |>
  mutate(
    `omega from ini()` = round(sqrt(`omega2 in ini()`), 4),
    # Reading A (correct): CV is the log-normal CV of the SD.
    `CV if lognormal` = round(100 * sqrt(exp(`omega2 in ini()`) - 1), 2),
    # Reading B (falsified): CV is simply omega * 100.
    `SD if CV = omega*100` = round(`CV printed` / 100, 4)
  )

knitr::kable(printed, caption = "Both readings of the C.V.(%) column, scored against the printed SD.")

# Reading A reproduces the printed CV column to every digit printed...
stopifnot(max(abs(printed$`CV if lognormal` - printed$`CV printed`)) < 0.01)
# ...and the omega implied by ini() rounds to the printed 2-dp SD column.
stopifnot(all(round(printed$`omega from ini()`, 2) == printed$`SD printed`))
# Reading B is FALSIFIED: it would require omega = 0.4311 against a printed
# SD of 0.41, and omega = 2.056 against a printed SD of 1.29.
stopifnot(max(abs(printed$`SD if CV = omega*100` - printed$`SD printed`)) > 0.5)
```

The `R.S.E.` column independently favours the higher-precision CV-derived
values: for CL, `0.062 / 0.4333 = 14.31%` against the printed 14.3%, whereas
`0.062 / 0.43 = 14.42%` would have printed as 14.4%.

# Typical-value profile

The model is loaded and its random effects zeroed to give the typical patient
at the model's own reference covariates (27.4 kg, eGFR 120 mL/min/1.73 m^2) and
at the cohort-median eGFR of 100.

```{r typical-profile}
mod  <- readModelDb("Marianski_2026_ceftriaxone")
modt <- rxode2::zeroRe(mod)

WT_REF   <- 27.4
GFR_REF  <- 120
GFR_MED  <- 100
MGKG     <- 50      # assumed regimen; see Assumptions and deviations
TINF     <- 0.5     # assumed 30-minute infusion

grid_24 <- seq(0, 24, by = 0.05)
ev24 <- as.data.frame(
  rxode2::et(amt = MGKG * WT_REF, cmt = "central", dur = TINF) |>
    rxode2::et(grid_24, cmt = "central")
)

typ <- bind_rows(
  rxode2::rxSolve(modt, ev24, params = c(WT = WT_REF, CRCL = GFR_REF),
                  returnType = "data.frame") |>
    mutate(scenario = "Reference eGFR 120"),
  rxode2::rxSolve(modt, ev24, params = c(WT = WT_REF, CRCL = GFR_MED),
                  returnType = "data.frame") |>
    mutate(scenario = "Cohort-median eGFR 100")
) |>
  filter(!is.na(Cc))

ggplot(typ, aes(time, Cc, colour = scenario)) +
  geom_line(linewidth = 0.8) +
  geom_hline(aes(yintercept = 10), linetype = "dashed", colour = "grey40") +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Whole-blood ceftriaxone (mg/L)", colour = NULL,
       title = "Typical 27.4 kg patient, 50 mg/kg IV over 30 min",
       subtitle = "Dashed line: total concentration whose 10% free fraction equals MIC 1 mg/L") +
  theme_bw() + theme(legend.position = "bottom")
```

At the cohort-median eGFR of 100 the typical patient's clearance is
`r round(0.91 * (100 / 120), 3)` L/h -- about 17% **below** the printed 0.91
L/h, because Table 1 normalises eGFR to 120 rather than to the cohort median.

# Gate 3: ODE solution against an exact matrix exponential

The model is a linear two-compartment system, so it has a closed-form solution.
Rather than transcribe a two-exponential formula (itself a transcription risk),
the exact solution is computed by eigendecomposition of the rate matrix and
compared against rxode2's integration using the **same drawn parameters**. Both
sides therefore share every parameter value and the only difference is
numerical, so a tight bound is correct here.

```{r gate-closed-form}
expmM <- function(M, t) {
  e <- eigen(M)
  e$vectors %*% diag(exp(e$values * t)) %*% solve(e$vectors)
}

# Exact amounts for a zero-order infusion of rate R0 over Tinf into `central`.
amt_exact <- function(kel, k12, k21, R0, Tinf, tt) {
  M    <- matrix(c(-(kel + k12), k12, k21, -k21), 2, 2)
  u    <- c(R0, 0)
  Minv <- solve(M)
  A_end <- Minv %*% (expmM(M, Tinf) - diag(2)) %*% u
  t(vapply(tt, function(tj) {
    if (tj <= Tinf) as.numeric(Minv %*% (expmM(M, tj) - diag(2)) %*% u)
    else            as.numeric(expmM(M, tj - Tinf) %*% A_end)
  }, numeric(2)))
}

rxode2::rxSetSeed(20260910)
set.seed(20260910)

n_cf <- 30L
cf_cov <- data.frame(id = 1:n_cf, WT = runif(n_cf, 6, 80), CRCL = runif(n_cf, 25, 225))
cf_ev <- bind_rows(lapply(1:n_cf, function(i) {
  d <- as.data.frame(
    rxode2::et(amt = MGKG * cf_cov$WT[i], cmt = "central", dur = TINF) |>
      rxode2::et(sort(unique(c(grid_24, TINF))), cmt = "central")
  )
  d$id <- i
  d
}))
cf_sim <- rxode2::rxSolve(mod, cf_ev, cf_cov, returnType = "data.frame", addCov = TRUE) |>
  filter(!is.na(Cc))

cf_dev <- vapply(split(cf_sim, cf_sim$id), function(z) {
  z  <- z[order(z$time), ]
  R0 <- MGKG * cf_cov$WT[cf_cov$id == z$id[1]] / TINF
  ex <- amt_exact(z$kel[1], z$k12[1], z$k21[1], R0, TINF, z$time)
  max(abs(ex[, 1] / z$vc[1] - z$Cc) / pmax(z$Cc, 1e-8))
}, numeric(1))

cat(sprintf("Closed form vs ODE, %d subjects: max relative deviation = %.2e\n",
            n_cf, max(cf_dev)))
stopifnot(max(cf_dev) < 1e-3)
```

This gate also validates the micro-constant algebra: `k12 = q / vc` and
`k21 = q / vp` and the two ODE right-hand sides all enter the comparison, so a
transposed volume in either would break it.

# Gate 4: mass balance holds at every time point

For a linear model eliminating only from the central compartment, the amount
eliminated by time `T` must equal the amount infused minus the amount still in
the body -- exactly, at any `T`, with no steady-state assumption:

`CL * AUC[0,T] == dose infused by T - central(T) - peripheral1(T)`

```{r gate-mass-balance}
mb_dev <- vapply(split(cf_sim, cf_sim$id), function(z) {
  z    <- z[order(z$time), ]
  dose <- MGKG * cf_cov$WT[cf_cov$id == z$id[1]]
  auc  <- cumsum(c(0, diff(z$time) * (head(z$Cc, -1) + tail(z$Cc, -1)) / 2))
  infused <- dose * pmin(z$time / TINF, 1)
  max(abs(z$cl[1] * auc - (infused - z$central - z$peripheral1)) / dose)
}, numeric(1))

cat(sprintf("Mass balance, %d subjects: max relative deviation = %.2e\n",
            n_cf, max(mb_dev)))
stopifnot(max(mb_dev) < 1e-3)
```

# Gate 5: the covariate exponents are recoverable from the built model

With the random effects zeroed, clearance and central volume are deterministic
functions of the covariates, so regressing the model's own evaluated `cl` and
`vc` on the supplied covariate table must return Table 1's exponents. The
left-hand side comes from rxode2's evaluation of `model()`; the right-hand side
only from the covariate table, so this is not circular -- a mis-typed exponent
in the model file changes the recovered coefficient.

```{r gate-exponents}
ex_cov <- data.frame(id = 1:40, WT = runif(40, 6, 80), CRCL = runif(40, 25, 225))
ex_ev <- bind_rows(lapply(1:40, function(i) {
  d <- as.data.frame(rxode2::et(amt = 1000, cmt = "central") |>
                       rxode2::et(c(1, 2), cmt = "central"))
  d$id <- i
  d
}))
ex_sim <- rxode2::rxSolve(modt, ex_ev, ex_cov, returnType = "data.frame", addCov = TRUE) |>
  filter(!is.na(Cc))

cl1 <- vapply(split(ex_sim$cl, ex_sim$id), function(x) x[1], numeric(1))
vc1 <- vapply(split(ex_sim$vc, ex_sim$id), function(x) x[1], numeric(1))
f_cl <- lm(log(cl1) ~ log(ex_cov$WT / 27.4) + log(ex_cov$CRCL / 120))
f_vc <- lm(log(vc1) ~ log(ex_cov$WT / 27.4))

recovered <- tibble::tibble(
  Quantity  = c("e_wt_cl", "eGFR exponent on CL", "CL_pop (L/h)",
                "e_wt_vc", "V1_pop (L)"),
  Published = c(0.75, 1, 0.91, 1, 8.22),
  Recovered = c(coef(f_cl)[2], coef(f_cl)[3], exp(coef(f_cl)[1]),
                coef(f_vc)[2], exp(coef(f_vc)[1]))
) |>
  mutate(Recovered = round(Recovered, 6),
         `Abs. difference` = signif(abs(Recovered - Published), 3))

knitr::kable(recovered, caption = "Table 1 covariate parameters recovered from the built model.")
stopifnot(max(recovered$`Abs. difference`) < 1e-6)
```

# PKNCA validation against a closed-form reference

The abstract reports **no** NCA parameters -- no Cmax, Tmax, AUC or half-life
appear anywhere in it -- so there is no published NCA table to compare against.
What Table 1 does supply is enough to derive the typical patient's NCA
parameters in closed form, which makes a genuine reference for the whole
simulate-then-integrate chain:

* `AUC(0,inf) = Dose / CL`, exactly;
* terminal half-life `= log(2) / lambda_z`, where `lambda_z` is the smaller
  eigenvalue of the rate matrix;
* `Cmax` and `Tmax` at the end of the infusion, from the matrix exponential.

```{r pknca}
nca_grid <- sort(unique(c(seq(0, 6, 0.02), seq(6.5, 48, 0.25), seq(49, 336, 1), TINF)))
DOSE_TYP <- MGKG * WT_REF
TRT <- "50 mg/kg IV over 0.5 h"

ev_nca <- as.data.frame(
  rxode2::et(amt = DOSE_TYP, cmt = "central", dur = TINF) |>
    rxode2::et(nca_grid, cmt = "central")
)
sim_nca_raw <- rxode2::rxSolve(modt, ev_nca, params = c(WT = WT_REF, CRCL = GFR_MED),
                               returnType = "data.frame") |>
  filter(!is.na(Cc))
if (is.null(sim_nca_raw$id)) sim_nca_raw$id <- 1L

# The far tail must not decay into solver noise, or PKNCA takes log() of a
# negative number and aucinf.obs becomes NaN.
stopifnot(all(sim_nca_raw$Cc >= 0), min(sim_nca_raw$Cc[sim_nca_raw$time > 0]) > 1e-4)

lam <- local({
  b <- sim_nca_raw$kel[1] + sim_nca_raw$k12[1] + sim_nca_raw$k21[1]
  (b - sqrt(b^2 - 4 * sim_nca_raw$kel[1] * sim_nca_raw$k21[1])) / 2
})
reference_nca <- tibble::tibble(
  treatment  = TRT,
  cmax       = max(sim_nca_raw$Cc),
  tmax       = TINF,
  aucinf.obs = DOSE_TYP / sim_nca_raw$cl[1],
  half.life  = log(2) / lam
)

sim_nca <- sim_nca_raw |>
  filter(!is.na(Cc)) |>
  transmute(id, time, Cc, treatment = TRT)
dose_nca <- data.frame(id = 1L, time = 0, amt = DOSE_TYP, treatment = TRT)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "mg/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_nca, amt ~ time | treatment + id, doseu = "mg")
intervals <- data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                        aucinf.obs = TRUE, half.life = TRUE, lambda.z = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r pknca-compare}
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = reference_nca,
  by            = "treatment",
  units         = c(cmax = "mg/L", tmax = "h",
                    aucinf.obs = "mg*h/L", half.life = "h"),
  tolerance_pct = 5
)
knitr::kable(cmp, caption = paste(
  "PKNCA on the simulated typical-value profile against the closed-form",
  "reference derived from Table 1. The reference is NOT a published NCA table;",
  "Marianski 2026 reports no NCA parameters. * differs by >5%."
), align = c("l", "l", "r", "r", "r"))

pct <- as.numeric(sub("%", "", sub("^\\+", "", cmp$`% diff`)))
stopifnot(max(abs(pct)) < 2)
```

All four parameters agree within 2%. The small half-life discrepancy is the
`lambda_z` regression picking up a little of the distribution phase; it is not
a model or transcription issue.

# The peripheral compartment is deep, and that matters

The published `Q` of 0.75 L/h against a `V2` of 13.74 L makes the peripheral
compartment equilibrate slowly, which gives this model a much longer terminal
half-life than ceftriaxone's literature value in children (roughly 5 to 9 h).

```{r terminal-phase}
terminal_half_life <- function(kel, k12, k21) {
  b <- kel + k12 + k21
  log(2) / ((b - sqrt(b^2 - 4 * kel * k21)) / 2)
}

th_typ <- terminal_half_life(sim_nca_raw$kel[1], sim_nca_raw$k12[1], sim_nca_raw$k21[1])
auc24  <- with(filter(sim_nca_raw, time <= 24),
               sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2))
auc_inf <- DOSE_TYP / sim_nca_raw$cl[1]

th_cohort <- vapply(split(cf_sim, cf_sim$id), function(z)
  terminal_half_life(z$kel[1], z$k12[1], z$k21[1]), numeric(1))

tibble::tibble(
  Quantity = c("Distribution half-life, typical (h)",
               "Terminal half-life, typical (h)",
               "Terminal half-life, cohort median (h)",
               "Terminal half-life, cohort 5th-95th pct (h)",
               "Share of AUC(0,inf) accrued in the first 24 h (%)"),
  Value = c(
    round(local({
      b <- sim_nca_raw$kel[1] + sim_nca_raw$k12[1] + sim_nca_raw$k21[1]
      log(2) / ((b + sqrt(b^2 - 4 * sim_nca_raw$kel[1] * sim_nca_raw$k21[1])) / 2)
    }), 2),
    round(th_typ, 1),
    round(median(th_cohort), 1),
    paste(round(quantile(th_cohort, c(0.05, 0.95)), 1), collapse = " to "),
    round(100 * auc24 / auc_inf, 1)
  )
) |>
  knitr::kable(caption = "Terminal-phase behaviour implied by the published Q and V2.")

# Structural, not stochastic: these follow from the published point estimates.
stopifnot(th_typ > 20, th_typ < 45)
stopifnot(100 * auc24 / auc_inf > 55, 100 * auc24 / auc_inf < 70)
```

Marianski 2026 sampled over 3 days and reported only **first-24-hour**
exposures, so the deep compartment is weakly identified -- `Q` and `V2` carry
the loosest precision in Table 1 (R.S.E. 26.3% and 33.2%) and the largest
interindividual variability in the model (CV 118% and 206%). Treat this model as
validated over the first day or two of therapy and do **not** use it to project
terminal elimination or long-interval accumulation. For context, the typical
patient accumulates about
`r round(local({ ev <- as.data.frame(rxode2::et(amt = DOSE_TYP, cmt = "central", dur = TINF, ii = 24, until = 24 * 20) |> rxode2::et(seq(0, 24 * 21, 0.25), cmt = "central")); d <- rxode2::rxSolve(modt, ev, params = c(WT = WT_REF, CRCL = GFR_MED), returnType = "data.frame") |> filter(!is.na(Cc)); d$Cc[which.min(abs(d$time - 24 * 20))] / d$Cc[which.min(abs(d$time - 24))] }), 2)`-fold
in trough concentration between the first dose and steady state on a q24h
schedule -- a prediction well outside what the source data can support.

# Virtual cohort

The abstract reports the age range (2 months to 17 years) and the eGFR
distribution (25 to 225, median 100 mL/min/1.73 m^2) but **not** the weight
distribution. Weights are therefore built from CDC median weight-for-age over
the reported age range, with lognormal scatter; eGFR is drawn lognormal with the
reported median and truncated to the reported range.

```{r cohort}
rxode2::rxSetSeed(20260910)
set.seed(20260910)

N_SUB <- 200L   # 200 per arm is the vignette cap

cdc_age <- c(0.167, 0.5, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10,
             11, 12, 13, 14, 15, 16, 17)
cdc_wt  <- c(5.6, 7.9, 10.2, 12.7, 14.3, 16.3, 18.5, 20.7, 23.0, 25.8, 28.7,
             32.0, 35.9, 40.5, 45.9, 51.0, 56.0, 60.5, 64.5)

cohort <- tibble::tibble(
  id   = seq_len(N_SUB),
  AGE  = runif(N_SUB, 0.167, 17),
  CRCL = pmin(pmax(100 * exp(rnorm(N_SUB, 0, 0.45)), 25), 225)
) |>
  mutate(WT = approx(cdc_age, cdc_wt, xout = AGE)$y * exp(rnorm(N_SUB, 0, 0.20)))

tibble::tibble(
  Covariate = c("Age (years)", "Weight (kg)", "eGFR (mL/min/1.73 m^2)"),
  `Simulated median` = round(c(median(cohort$AGE), median(cohort$WT),
                               median(cohort$CRCL)), 1),
  `Simulated range` = c(
    paste(round(range(cohort$AGE), 2), collapse = " to "),
    paste(round(range(cohort$WT), 1), collapse = " to "),
    paste(round(range(cohort$CRCL), 0), collapse = " to ")
  ),
  `Reported by Marianski 2026` = c("2 months to 17 years", "not reported",
                                   "25 to 225, median 100")
) |>
  knitr::kable(caption = "Virtual cohort against the reported population.")
```

The simulated median weight lands near the 27.4 kg that Table 1 uses to
normalise its allometric terms, which is weak corroboration that 27.4 kg was
the cohort median -- the abstract never says so.

```{r cohort-sim}
cohort_cov <- as.data.frame(cohort[, c("id", "WT", "CRCL")])
ev_cohort <- bind_rows(lapply(seq_len(N_SUB), function(i) {
  d <- as.data.frame(
    rxode2::et(amt = MGKG * cohort$WT[i], cmt = "central", dur = TINF) |>
      rxode2::et(seq(0, 24, by = 0.1), cmt = "central")
  )
  d$id <- i
  d
}))
sim_cohort <- rxode2::rxSolve(mod, ev_cohort, cohort_cov,
                              returnType = "data.frame", addCov = TRUE) |>
  filter(!is.na(Cc))

stopifnot(all(sim_cohort$Cc >= 0), !anyNA(sim_cohort$Cc),
          length(unique(sim_cohort$id)) == N_SUB)

sim_cohort |>
  group_by(time) |>
  summarise(p05 = quantile(Cc, 0.05), p50 = median(Cc),
            p95 = quantile(Cc, 0.95), .groups = "drop") |>
  ggplot(aes(time)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = p50), colour = "steelblue", linewidth = 0.9) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Whole-blood ceftriaxone (mg/L)",
       title = "Simulated cohort, 50 mg/kg IV over 30 min",
       subtitle = "Median with 5th-95th percentile band, 200 subjects") +
  theme_bw()
```

# Replicating Figure 2: free time over MIC

Marianski 2026 computed the fraction of the **first 24 hours** during which the
free ceftriaxone concentration exceeded the MIC, applying a literature free
fraction of 10%, for MICs of 1 (susceptible), 2 (intermediate), and 4 and 8
mg/L (resistant).

```{r ftmic}
FREE_FRACTION <- 0.10   # Methods; Figure 2 caption ("90% protein binding")
MICS <- c(1, 2, 4, 8)

ftmic <- bind_rows(lapply(MICS, function(mic) {
  tibble::tibble(
    MIC = mic,
    id  = as.integer(names(split(sim_cohort$Cc, sim_cohort$id))),
    fT  = vapply(split(sim_cohort$Cc, sim_cohort$id),
                 function(cc) 100 * mean(FREE_FRACTION * cc > mic), numeric(1))
  )
}))

ggplot(ftmic, aes(fT)) +
  geom_density(aes(fill = factor(MIC)), alpha = 0.6, colour = "grey20",
               bounds = c(0, 100)) +
  geom_rug(sides = "b", alpha = 0.5) +
  geom_vline(xintercept = 50, linetype = "dashed") +
  facet_wrap(~ MIC, ncol = 1, strip.position = "left",
             labeller = labeller(MIC = function(x) paste(x, "mg/L"))) +
  scale_fill_viridis_d(option = "D", direction = -1, guide = "none") +
  labs(x = "Free time over MIC, percent of the first 24 h", y = "MIC",
       title = "Replicates Figure 2 of Marianski 2026",
       subtitle = "Dashed line: the paper's 50% fT>MIC reference") +
  theme_bw()
```

```{r ftmic-table}
ft_summary <- ftmic |>
  group_by(MIC) |>
  summarise(
    `5th pct`  = round(quantile(fT, 0.05), 1),
    `1st quartile` = round(quantile(fT, 0.25), 1),
    Median     = round(median(fT), 1),
    `3rd quartile` = round(quantile(fT, 0.75), 1),
    `95th pct` = round(quantile(fT, 0.95), 1),
    `% of subjects at or above 50%` = round(100 * mean(fT >= 50), 1),
    .groups = "drop"
  )
knitr::kable(ft_summary, caption = "Simulated fT>MIC distribution by MIC.")
```

The paper states three claims about this figure. Each is checked below on the
**centre and robust quantiles** of the simulated cohort, never on its extremes:
the extreme of a random cohort is not reproducible across rxode2 versions or
solver thread counts, whereas a median or interquartile width is.

```{r ftmic-claims}
med <- function(mic) ft_summary$Median[ft_summary$MIC == mic]
iqrw <- function(mic) with(ft_summary[ft_summary$MIC == mic, ],
                           `3rd quartile` - `1st quartile`)

claims <- tibble::tribble(
  ~Claim, ~`Paper (Results)`, ~Achieved, ~Pass,
  "fT>MIC exceeds 50% at MIC 1 mg/L",
    "\"fT>MIC was >50% at CRO MICs <=1 mg/L\"",
    paste0("median ", med(1), "%"),
    med(1) > 50,
  "fT>MIC is variable at MIC 2 mg/L",
    "\"variable for CRO MICs of 2-4 mg/L\"",
    paste0("interquartile width ", iqrw(2), " pp"),
    iqrw(2) > 20,
  "fT>MIC is variable at MIC 4 mg/L",
    "\"variable for CRO MICs of 2-4 mg/L\"",
    paste0("interquartile width ", iqrw(4), " pp"),
    iqrw(4) > 20,
  "fT>MIC falls below 50% at MIC 8 mg/L",
    "\"< 50% at 8 mg/L\"",
    paste0("median ", med(8), "%"),
    med(8) < 50
)
knitr::kable(claims, caption = "The paper's three printed fT>MIC claims, scored against the simulated cohort.")
stopifnot(all(claims$Pass))
```

All three reproduce. The typical patient alone gives
`r round(100 * mean(FREE_FRACTION * filter(typ, scenario == "Cohort-median eGFR 100")$Cc > 1), 1)`%,
`r round(100 * mean(FREE_FRACTION * filter(typ, scenario == "Cohort-median eGFR 100")$Cc > 2), 1)`%,
`r round(100 * mean(FREE_FRACTION * filter(typ, scenario == "Cohort-median eGFR 100")$Cc > 4), 1)`% and
`r round(100 * mean(FREE_FRACTION * filter(typ, scenario == "Cohort-median eGFR 100")$Cc > 8), 1)`%
at MICs of 1, 2, 4 and 8 mg/L, which straddles 50% between MIC 2 and MIC 4
exactly as the paper's "variable for 2-4 mg/L" describes, and supports the
Conclusion that "therapeutic drug monitoring approaches ... are needed for all
MICs >=2 mg/L".

The **shape** of the simulated MIC-2 density sits higher than the published
Figure 2 panel, whose visual mode is near 55-60%. The dosing regimen is not
reported (see below), so the level of the fT>MIC distribution is not fully
determined by the paper; only the three printed statements above are, and those
are what the gate tests.

# Renal function drives exposure

Because eGFR enters clearance as a through-origin linear ratio, exposure scales
inversely with renal function over the whole reported range.

```{r renal-strata}
strata <- tibble::tibble(
  stratum = c("eGFR 30 (impaired)", "eGFR 60", "eGFR 100 (cohort median)",
              "eGFR 150", "eGFR 220 (hyperfiltration)"),
  CRCL    = c(30, 60, 100, 150, 220)
)
strat_sim <- bind_rows(lapply(seq_len(nrow(strata)), function(i) {
  rxode2::rxSolve(modt, ev24, params = c(WT = WT_REF, CRCL = strata$CRCL[i]),
                  returnType = "data.frame") |>
    filter(!is.na(Cc)) |>
    mutate(stratum = strata$stratum[i], CRCL = strata$CRCL[i])
}))

strat_raw <- strat_sim |>
  group_by(stratum, CRCL) |>
  summarise(
    cl_exact = first(cl),
    auc24    = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    ft1      = 100 * mean(FREE_FRACTION * Cc > 1),
    ft4      = 100 * mean(FREE_FRACTION * Cc > 4),
    .groups  = "drop"
  ) |>
  arrange(CRCL)

strat_tbl <- strat_raw |>
  transmute(stratum, CRCL,
            `CL (L/h)` = round(cl_exact, 3),
            `AUC(0,24) (mg*h/L)` = round(auc24, 0),
            `fT>MIC at 1 mg/L (%)` = round(ft1, 1),
            `fT>MIC at 4 mg/L (%)` = round(ft4, 1))
knitr::kable(strat_tbl, caption = "Typical 27.4 kg patient across the reported eGFR range.")

# Structural and deterministic: CL is exactly proportional to eGFR. Assert on
# the UNROUNDED clearances -- the display column above is rounded to 3 decimals,
# and a ratio of rounded values is not exact.
stopifnot(all(diff(strat_raw$cl_exact) > 0))
stopifnot(abs(strat_raw$cl_exact[strat_raw$CRCL == 60] /
              strat_raw$cl_exact[strat_raw$CRCL == 30] - 2) < 1e-8)
# The whole term is linear through the origin, so CL / eGFR is constant.
stopifnot(diff(range(strat_raw$cl_exact / strat_raw$CRCL)) < 1e-12)
```

A child at the top of the reported eGFR range clears ceftriaxone about
`r round(220 / 30, 1)`-fold faster than one at the bottom, and this term alone
moves fT>MIC at 4 mg/L from
`r strat_tbl$"fT>MIC at 4 mg/L (%)"[strat_tbl$CRCL == 30]`% to
`r strat_tbl$"fT>MIC at 4 mg/L (%)"[strat_tbl$CRCL == 220]`% -- which is the
paper's point that glomerular filtration was "significantly associated but not
fully predictive" and that variability persists after accounting for it.

# Assumptions and deviations

**Values taken from a raster image.** Every parameter comes from Table 1, which
is published only as an embedded JPEG. The values were read from the decoded
image, not from a text layer. Gate 1 above is the check on that read: all nine
rows satisfy `R.S.E. = S.E. / Value` to within the printing precision, which a
mis-read digit would break. A transcription of all three panels is stored beside
the source PDF as `unknown_2026_PMC12792806_panels_transcribed.md`.

**Conference abstract, not a peer-reviewed paper.** This is IDWeek 2025 poster
P-1254. A conference abstract's covariate model should be treated as
provisional: the cefepime arm of this same PALISI/VAMS study (Antimicrob Agents
Chemother 2026, PMC13321836) retained a different renal-covariate form. A
Europe PMC search on the funding grant R01HD103755 returns this abstract plus
that cefepime paper and no full-text ceftriaxone publication, so as of the
extraction date there is nothing fuller to supersede it.

**Concentrations are whole blood, not plasma.** The model was fitted to
whole-blood ceftriaxone from VAMS microsamples, so `Cc` is a whole-blood
concentration and is not interchangeable with the plasma concentrations that
other ceftriaxone popPK models report, nor directly with plasma-referenced
susceptibility breakpoints. The paper's own fT>MIC analysis applies a
literature **plasma** protein-binding fraction of 90% directly to these
whole-blood concentrations; that approximation is the authors' and is
reproduced here unchanged. The same group has published a whole-blood-to-plasma
translation for VAMS antibiotic assays (Ther Drug Monit 2026, PMC13366314), but
this abstract neither cites it nor states that any translation was applied
before modelling.

**Dosing regimen assumed.** The abstract never states a dose, frequency or
infusion duration -- only that patients were "prescribed CRO" as standard of
care and that exposures were computed "from exact dosing and covariate
histories". This vignette uses **50 mg/kg as a 30-minute infusion**, the most
common paediatric ceftriaxone regimen. All fT>MIC levels shown here are
conditional on that choice; the three claims tested in the gate are the paper's
own printed statements, which the assumed regimen reproduces.

**Weight distribution assumed.** No cohort weight distribution is reported.
Weights are CDC median weight-for-age over the reported 2-month-to-17-year age
range with 20% lognormal scatter, and ages are drawn uniformly because the
abstract gives only the range. The resulting median weight lands near Table 1's
27.4 kg normalisation constant, which is consistent with -- but not proof of --
27.4 kg being the cohort median.

**eGFR distribution assumed.** Drawn lognormal with the reported median of 100
and truncated to the reported 25-225 range. The abstract reports only the range
and median, not the shape.

**The eGFR reference is 120, not the cohort median.** Table 1 normalises eGFR to
120 mL/min/1.73 m^2 while Results gives the cohort median as 100, so the printed
CL of 0.91 L/h is the clearance of a 27.4 kg child with **normal** renal
function, about 17% above the typical patient in this cohort. This is unusual --
most models in this library centre on the cohort median -- and is reproduced as
printed.

**No covariates on Q or V2.** Table 1 prints `Q_i = Q_pop` and `V2_i = V2_pop`
with no weight term, which is unusual for an allometrically scaled model. It is
encoded faithfully; a downstream user scaling this model to a very different
body size should note that only CL and V1 respond to weight.

**Terminal phase is not validated.** As quantified above, the published Q and V2
imply a typical terminal half-life near 30 h against a literature value of
roughly 5 to 9 h in children, and a 2.25-fold trough accumulation on q24h
dosing. The study sampled 3 days and reported only first-24-hour exposures, and
Q and V2 are the least precisely estimated parameters in Table 1. Use this model
for early-therapy exposure, not for terminal elimination or accumulation.

**No NCA comparison against the paper.** The abstract reports no NCA
parameters, so the comparison table above scores PKNCA against a closed-form
reference derived from Table 1, not against published values. This validates the
simulate-then-integrate chain; it is not an external check on the model.

**Residual error is proportional only.** Table 1's error-model block has a `b`
row and no `a` row, which in Monolix identifies a pure proportional error model.
No additive term is included.

**Interindividual variability on Q and V2 is enormous.** CV 118% and 206%
respectively. Both are reported with usable precision (R.S.E. 24.3% and 19.3%)
and are carried as published, but they make individual peripheral-compartment
predictions nearly uninformative and produce the very wide terminal half-life
spread shown above.
