---
title: "Oral salbutamol (Marques 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Oral salbutamol (Marques 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Marques_2025_salbutamol"))$reference`
* Article: <https://doi.org/10.3390/pharmaceutics17010039>
* Supplementary material: <https://www.mdpi.com/article/10.3390/pharmaceutics17010039/s1>

Marques and Vale built a GastroPlus PBPK model of a 4 mg oral salbutamol
immediate-release tablet, used its Population Simulator to generate 40 virtual
patients, extracted 838 simulated plasma concentrations over 24 h of 4 mg q6h
dosing, and fitted a population PK model to that synthetic dataset in Monolix
2024R1. There are no clinical observations anywhere in the analysis. The
reported final model is two-compartment with transit-compartment first-order
absorption and linear elimination; its parameter estimates are the paper's
Table 6.

**This package entry encodes Table 6 exactly as printed.** That is a
deliberate choice, and it needs a warning attached: the Table 6 estimates do
not reproduce the paper's own published summaries of the same fit. Simulating
them gives a profile that peaks late in the dosing interval and is almost flat,
against a published VPC that peaks about an hour after each dose with a roughly
three-fold peak-to-trough swing, and it underpredicts the paper's own
Supplementary Table S3 exposure by five- to six-fold. The paper's Table 7
reports empirical-Bayes geometric means for three of the same parameters that
differ from Table 6 by factors of 32 to 45 (Mtt), about 2 (Cl) and about 3.4
(V1).

Everything below quantifies those discrepancies rather than smoothing them
over. They are recorded here as **errata in the source publication**, not as
evidence that the model is fabricated: the fit itself is plausible, and the
most economical explanation for the Mtt gap is a unit or transcription error in
the table. A correspondence to the authors is the appropriate route to
resolution, and until it is resolved the packaged model is the published table.
The "Errata" section at the end lists every discrepancy found, and
"What this model is not" explains why no corrected variant is shipped.

```{r model-desc, echo = FALSE, results = "asis"}
cat(rxode2::rxode2(readModelDb("Marques_2025_salbutamol"))$description)
```

# Population

The population is the 40-patient virtual cohort of Marques 2025 Table 5:
age 5-65 years (mean 39.5, SD 18.4), weight 26-109 kg (mean 60.2, SD 19.8),
BSA 0.94-2.37 m^2 (mean 1.64, SD 0.330), 16 female (40%) and 24 male (60%),
race American 20 (50%) / Asian 10 (25%) / Chinese 10 (25%), and health status
healthy 26 (65%) / obese 4 (10%) / cirrhosis Child-Pugh A 10 (25%). CYP2C19
and CYP2D6 relative expression averaged 1.30% (SD 1.50) and 1.90% (SD 1.60).
Per-group demographics are in Supplementary Table S2.

These are simulated physiologies from the GastroPlus PEAR age-related
population generator, not measured subjects. The cohort simulated below is
therefore a cohort of a cohort: virtual patients drawn from a model fitted to
virtual patients.

```{r population-meta}
local({
  fbody <- body(readModelDb("Marques_2025_salbutamol"))
  meta_env <- new.env()
  for (stmt in as.list(fbody)[-1]) {
    if (is.call(stmt) && length(stmt) >= 1 &&
        identical(stmt[[1]], as.name("<-"))) {
      eval(stmt, envir = meta_env)
    } else {
      break
    }
  }
  cat("Population:\n")
  str(meta_env$population, max.level = 1)
  cat("\nCovariates screened by the authors but NOT encodable:\n")
  str(meta_env$covariatesDataExcluded, max.level = 1)
})
```

# Source trace

Every structural equation and every `ini()` value, with its source location.

| Model element | Value | Source location |
|---|---|---|
| Structure: 2-cmt, transit absorption, linear elimination | n/a | Section 3.3.1 paragraph 1; Abstract; Discussion paragraph 2; Figure 5 caption |
| `lmtt` (mean transit time Mtt) | 9.50 h (RSE 7.14%) | Table 6, Fixed effects |
| `lktr` (transit rate Ktr) | 0.15 1/h (RSE 9.21%) | Table 6, Fixed effects |
| `nn` (transit chain length) | 0.425, derived as Mtt*Ktr - 1 | Monolix transit parameterisation; both Table 6 values reproduced exactly |
| `lka` (absorption rate ka) | 2.91 1/h (RSE 6.80%) | Table 6, Fixed effects |
| `lcl` (apparent clearance Cl/F) | 140 L/h (RSE 223%) | Table 6, Fixed effects |
| `lvc` (apparent central volume V1/F) | 77.5 L (RSE 97.2%) | Table 6, Fixed effects |
| `lq` (apparent intercompartmental clearance Q/F) | 48.9 L/h (RSE 24.9%) | Table 6, Fixed effects |
| `lvp` (apparent peripheral volume V2/F) | 130 L (RSE 21.4%) | Table 6, Fixed effects |
| `etalmtt` | 0.43^2 = 0.1849 | Table 6, Random effects IIV(Mtt), RSE 13.0% |
| `etalktr` | 0.32^2 = 0.1024 | Table 6, Random effects IIV(Ktr), RSE 13.8% |
| `etalka` | 0.41^2 = 0.1681 | Table 6, Random effects IIV(ka), RSE 15.4% |
| `etalcl` | 0.38^2 = 0.1444 | Table 6, Random effects IIV(Cl), RSE 13.0% |
| `etalvc` | 0.36^2 = 0.1296 | Table 6, Random effects IIV(V1), RSE 12.3% |
| `etalq` | 0.92^2 = 0.8464 | Table 6, Random effects IIV(Q), RSE 14.9% |
| `etalvp` | 0.87^2 = 0.7569 | Table 6, Random effects IIV(V2), RSE 27.4% |
| Diagonal omega (no correlations) | n/a | Section 3.3.2: "No significant correlations were found between PK parameters" |
| `addSd` | 0.19e-4 = 1.9e-5 ug/mL (RSE 20.2%) | Table 6, Error model parameters, `a` |
| `propSd` | 0.15e-2 = 1.5e-3 (RSE 37.2%) | Table 6, Error model parameters, `b` |
| `combined1()` (SDs add linearly) | n/a | Section 3.3.1 error equation Y = F + (a + b*F)*eps |
| Log-normal IIV (omegas are log-scale SDs) | n/a | Inferred from Table 7 geometric SDs; see Errata item 6 |
| Dose regimen 4 mg q6h | n/a | Section 2.2; Table S3 caption |
| Reference NCA, single 4 mg dose | Cmax 0.00725 / 0.00919 ug/mL | Table 4, Observed / Predicted rows |
| Reference NCA, 4 mg q6h | Cmax 0.0125-0.0151 ug/mL | Table S3, groups 1-4 |
| Model selection criteria | BICc / -2LL | Table S4 (see Errata item 4) |
| Covariate screen | correlations and p-values | Table S5 (see Errata item 5) |

# Model handle and a required rxode2 workaround

The packaged model uses rxode2's built-in `transit()`, the closed-form
(gamma-density) input rate of the Savic transit-absorption model. On the
rxode2 version installed here, `transit()` silently evaluates to zero when the
model is solved through an `rxUi` object obtained from `readModelDb()`, while
the *identical* generated ODE code solves correctly as a plain `rxode2` model.
The check below documents the defect and builds the plain-ODE twin used for
every simulation in this vignette. All parameter values are read out of the
packaged model's `iniDf`, so nothing is re-typed from the paper here.

```{r model-handle}
mod <- rxode2::rxode2(readModelDb("Marques_2025_salbutamol"))
ini <- mod$iniDf

# Plain-ODE twin: same generated code, solved outside the rxUi wrapper.
mod_ode <- rxode2::rxode2(rxode2::rxNorm(mod))

theta_names <- ini$name[!is.na(ini$ntheta)]
eta_names   <- ini$name[!is.na(ini$neta1)]

# Structural fixed effects only; the residual-error terms are not ODE inputs.
theta <- setNames(ini$est[!is.na(ini$ntheta)], theta_names)
theta <- theta[!names(theta) %in% c("addSd", "propSd")]
eta_var <- setNames(ini$est[!is.na(ini$neta1)], eta_names)

# Demonstrate the defect: same code, two handles, one of them returns zero.
ev_probe <- rbind(
  data.frame(id = 1L, time = 0, amt = 4, evid = 1L, cmt = "depot"),
  data.frame(id = 1L, time = seq(0, 24, by = 0.5), amt = 0, evid = 0L,
             cmt = "central")
)
ev_probe <- ev_probe[order(ev_probe$time, -ev_probe$evid), ]
ev_probe_eta <- ev_probe
for (nm in eta_names) ev_probe_eta[[nm]] <- 0

cmax_ui  <- max(suppressWarnings(rxode2::rxSolve(
  rxode2::zeroRe(mod), ev_probe, returnType = "data.frame"))$Cc, na.rm = TRUE)
cmax_ode <- max(rxode2::rxSolve(
  mod_ode, ev_probe_eta, params = theta, returnType = "data.frame")$Cc)

tibble::tibble(
  Handle = c("rxUi from readModelDb() (transit() returns 0)",
             "plain rxode2 model from rxNorm() of the same rxUi"),
  `Cmax (ug/mL)` = c(cmax_ui, cmax_ode)
) |>
  knitr::kable(digits = 7, caption = "The rxode2 transit()/rxUi defect that forces the plain-ODE twin.")

# Lock the workaround in: if a future rxode2 fixes the rxUi path, cmax_ui
# becomes non-zero and this assertion fires, prompting the simplification.
stopifnot(cmax_ui == 0, cmax_ode > 0)
```

## Transit-chain self-consistency

Table 6 prints Mtt and Ktr as two separate estimates. Monolix's
transit-absorption parameterisation derives the (continuous) number of transit
compartments as `n = Mtt * Ktr - 1`, and rxode2's `transit(n, mtt)` uses
`ktr = (n + 1) / mtt` internally. The two printed numbers are therefore
mutually consistent, and this encoding reproduces both of them exactly.

```{r transit-check}
sim_probe <- rxode2::rxSolve(mod_ode, ev_probe_eta, params = theta,
                             returnType = "data.frame")

nn_typ  <- sim_probe$nn[1]
mtt_typ <- sim_probe$mtt[1]
ktr_typ <- sim_probe$ktr[1]

tibble::tibble(
  Quantity = c("Mtt (h)", "Ktr (1/h), Table 6 value",
               "n = Mtt*Ktr - 1", "Ktr recovered as (n+1)/Mtt"),
  Value = c(mtt_typ, ktr_typ, nn_typ, (nn_typ + 1) / mtt_typ)
) |>
  knitr::kable(digits = 6, caption = "Table 6's Mtt/Ktr pair is internally self-consistent and exactly reproduced.")

stopifnot(
  all.equal(mtt_typ, 9.50),
  all.equal(ktr_typ, 0.15),
  all.equal(nn_typ, 0.425),
  all.equal((nn_typ + 1) / mtt_typ, 0.15)
)
```

## Mass-balance gate (single dose)

For a single dose the closed-form transit chain delivers the whole dose, so
AUC extrapolated to infinity must equal `Dose / (CL/F)` exactly. This is the
strongest available check that the ODE system, the bioavailability suppression
`f(depot) <- 0`, and the transit input are all wired up correctly, and it is
independent of anything the paper reports.

```{r mass-balance}
solve_typ <- function(dose_times, t_end, by = 0.01) {
  grid <- sort(unique(c(dose_times, seq(0, t_end, by = by))))
  ev <- dplyr::bind_rows(
    tibble::tibble(id = 1L, time = dose_times, amt = 4, evid = 1L,
                   cmt = "depot"),
    tibble::tibble(id = 1L, time = grid, amt = 0, evid = 0L, cmt = "central")
  ) |>
    dplyr::arrange(time, dplyr::desc(evid)) |>
    as.data.frame()
  for (nm in eta_names) ev[[nm]] <- 0
  rxode2::rxSolve(mod_ode, ev, params = theta, returnType = "data.frame")
}

trapz <- function(x, y) sum(diff(x) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)

sd_long <- solve_typ(0, 300, by = 0.01)
auc_sd  <- trapz(sd_long$time, sd_long$Cc)
cl_typ  <- exp(theta[["lcl"]])

tibble::tibble(
  Quantity = c("Simulated AUC(0-300 h), single 4 mg dose (ug*h/mL)",
               "Dose / (CL/F) = 4 mg / 140 L/h (ug*h/mL)",
               "Ratio"),
  Value = c(auc_sd, 4 / cl_typ, auc_sd / (4 / cl_typ))
) |>
  knitr::kable(digits = 7, caption = "Single-dose mass balance is exact: the ODE encoding is correct.")

stopifnot(abs(auc_sd / (4 / cl_typ) - 1) < 1e-4)
```

## Dose carryover under q6h dosing

`transit()` is a closed form driven by `tad()` and `podo()` -- time since, and
amount of, the **most recent** dose. Each new dose therefore restarts the
transit input and discards whatever remains un-absorbed from the previous one.
That approximation is harmless when the mean transit time is short relative to
the dosing interval, but Table 6's Mtt of 9.50 h is *longer* than the 6 h
interval, so a substantial fraction of each dose is discarded.

This is a property of the closed-form transit implementation, not a defect in
the paper. It is quantified here because it affects every multiple-dose number
below, and because it is one more reason an Mtt of 9.50 h is hard to reconcile
with a q6h dataset.

```{r carryover}
auc_q6  <- trapz(solve_typ(c(0, 6, 12, 18), 400, by = 0.02)$time,
                 solve_typ(c(0, 6, 12, 18), 400, by = 0.02)$Cc)
auc_q24 <- trapz(solve_typ(c(0, 24, 48, 72), 400, by = 0.02)$time,
                 solve_typ(c(0, 24, 48, 72), 400, by = 0.02)$Cc)

tibble::tibble(
  Regimen = c("4 x 4 mg q6h (the paper's regimen)",
              "4 x 4 mg q24h (interval >> Mtt, for contrast)"),
  `Simulated AUCinf (ug*h/mL)` = c(auc_q6, auc_q24),
  `Total dose / (CL/F)` = 16 / cl_typ,
  `Fraction delivered` = c(auc_q6, auc_q24) / (16 / cl_typ)
) |>
  knitr::kable(digits = 5, caption = "Closed-form transit discards un-absorbed drug at each new dose; the loss is large when Mtt (9.50 h) exceeds the dosing interval (6 h).")
```

# Typical-value profiles

```{r typical-profiles}
typ_sd <- solve_typ(0, 24, by = 0.01)
typ_md <- solve_typ(c(0, 6, 12, 18), 24, by = 0.01)

last_int <- typ_md |> dplyr::filter(time >= 18)

typ_stats <- tibble::tibble(
  Quantity = c("Single dose: Cmax (ug/mL)", "Single dose: Tmax (h)",
               "q6h 0-24 h: Cmax (ug/mL)", "q6h 0-24 h: Tmax (h)",
               "q6h 0-24 h: AUC (ug*h/mL)",
               "q6h last interval: Tmax after dose (h)",
               "q6h last interval: peak/trough ratio"),
  Value = c(max(typ_sd$Cc), typ_sd$time[which.max(typ_sd$Cc)],
            max(typ_md$Cc), typ_md$time[which.max(typ_md$Cc)],
            trapz(typ_md$time, typ_md$Cc),
            last_int$time[which.max(last_int$Cc)] - 18,
            max(last_int$Cc) / min(last_int$Cc))
)
knitr::kable(typ_stats, digits = 5, caption = "Typical-value behaviour of Table 6 as printed.")
```

```{r figure-2-typical, fig.cap = "Typical-value profiles from Table 6 as printed. Compare with Figure 2 of Marques 2025, whose panel (a) single-dose profile peaks at about 1.3 h at 0.0133 ug/mL and whose panel (b) q6h profile shows four sharp peaks. The Table 6 profile instead peaks late and is nearly flat under q6h dosing."}
dplyr::bind_rows(
  typ_sd |> dplyr::transmute(time, Cc, Regimen = "Single 4 mg dose"),
  typ_md |> dplyr::transmute(time, Cc, Regimen = "4 mg q6h")
) |>
  ggplot(aes(time, Cc, colour = Regimen)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0.0133, linetype = "dashed", colour = "grey40") +
  annotate("text", x = 12, y = 0.0142, label = "Figure 2 PBPK Cmax = 0.0133 ug/mL",
           size = 3, colour = "grey30") +
  scale_x_continuous(breaks = seq(0, 24, by = 6)) +
  labs(x = "Time (h)", y = "Plasma salbutamol (ug/mL)",
       title = "Table 6 as printed, typical values") +
  theme_minimal()
```

# Virtual cohort and simulation

The model carries no covariates, so the cohort is defined entirely by the seven
inter-individual random effects of Table 6. We simulate 200 subjects per arm --
ample for a prediction interval and well inside the vignette time budget --
under both the single-dose and the q6h regimens. Random effects are drawn once
and applied to both arms, so the two arms differ only by regimen.

```{r cohort}
set.seed(2026L)
n_sub <- 200L
dose_times <- c(0, 6, 12, 18)
grid_times <- sort(unique(c(dose_times, seq(0, 24, by = 0.1))))

etas <- as.data.frame(lapply(
  eta_var, function(v) stats::rnorm(n_sub, mean = 0, sd = sqrt(v))
))
etas$id <- seq_len(n_sub)

arm_events <- function(dts, label) {
  dplyr::bind_rows(
    tidyr::expand_grid(id = seq_len(n_sub), time = dts) |>
      dplyr::mutate(amt = 4, evid = 1L, cmt = "depot"),
    tidyr::expand_grid(id = seq_len(n_sub), time = grid_times) |>
      dplyr::mutate(amt = 0, evid = 0L, cmt = "central")
  ) |>
    dplyr::mutate(treatment = label)
}

solve_arm <- function(dts, label) {
  ev <- arm_events(dts, label) |>
    dplyr::left_join(etas, by = "id") |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    as.data.frame()
  suppressWarnings(
    rxode2::rxSolve(mod_ode, ev, params = theta, keep = "treatment",
                    returnType = "data.frame")
  )
}

sim <- dplyr::bind_rows(
  solve_arm(0, "Single 4 mg dose"),
  solve_arm(dose_times, "4 mg q6h")
)

nrow(sim)
```

A quarter of the cohort has a **negative** implied transit-chain length. Table 6
puts IIV on both Mtt and Ktr, so `n = Mtt * Ktr - 1` varies between subjects and
goes negative whenever `Mtt * Ktr < 1`; the published omegas make that true for
roughly 23% of draws. rxode2 still evaluates the gamma-form input (the
requirement is only `n > -1`, which always holds), but a negative number of
transit compartments has no physical interpretation, and the most extreme draws
are numerically stiff enough that a small number of subjects fail to solve.

```{r negative-n}
subj_diag <- sim |>
  dplyr::group_by(treatment, id) |>
  dplyr::summarise(nn = dplyr::first(nn), failed = any(is.na(Cc)),
                   .groups = "drop")

diag_tbl <- subj_diag |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    `Subjects` = dplyr::n(),
    `n < 0 (%)` = 100 * mean(nn < 0),
    `Minimum n` = min(nn),
    `Failed to solve` = sum(failed),
    .groups = "drop"
  ) |>
  dplyr::rename(Regimen = treatment)

knitr::kable(diag_tbl, digits = 3, caption = "Consequence of Table 6 putting IIV on both Mtt and Ktr: a negative implied transit-chain length for about a quarter of the population.")

# 47 of 200 draws are negative; exactly one q6h subject fails to solve.
stopifnot(
  sum(subj_diag$nn < 0) == 94L,           # 47 subjects x 2 arms (shared etas)
  sum(subj_diag$failed) == 1L
)

# Drop any subject that did not solve, in either arm, so both arms carry the
# same subjects and the NCA below is not silently unbalanced.
bad_ids <- sort(unique(subj_diag$id[subj_diag$failed]))
sim <- dplyr::filter(sim, !id %in% bad_ids)
n_kept <- dplyr::n_distinct(sim$id)
n_kept
```

```{r figure-5-vpc, fig.cap = "Simulated q6h prediction interval from Table 6 as printed (median and 90% interval, N = 199). This is the analogue of Figure 5 of Marques 2025. The published VPC spans a single 0-6 h dosing interval and shows a peak about 1 h after the dose with a roughly three-fold peak-to-trough fall; the simulated profile instead rises slowly across the whole 24 h."}
sim |>
  dplyr::filter(treatment == "4 mg q6h", !is.na(Cc), time > 0) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    med = stats::median(Cc),
    lo  = stats::quantile(Cc, 0.05),
    hi  = stats::quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, med)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.18, fill = "steelblue") +
  geom_line(colour = "firebrick", linewidth = 0.8) +
  geom_vline(xintercept = dose_times, linetype = "dotted", colour = "grey60") +
  scale_x_continuous(breaks = seq(0, 24, by = 6)) +
  labs(x = "Time (h)", y = "Plasma salbutamol (ug/mL)",
       title = "4 mg oral salbutamol q6h, Table 6 as printed",
       caption = "Red = median, ribbon = 90% prediction interval. Dotted lines = dose times.") +
  theme_minimal()
```

# PKNCA validation

NCA is run per subject and per arm with PKNCA over the 0-24 h window the paper
simulated. The time-zero record is retained deliberately: dropping it makes
PKNCA warn that the AUC interval starts before the first measurement.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

stopifnot(0 %in% sim_nca$time)

dose_df <- dplyr::bind_rows(
  tidyr::expand_grid(id = unique(sim_nca$id), time = 0) |>
    dplyr::mutate(amt = 4, treatment = "Single 4 mg dose"),
  tidyr::expand_grid(id = unique(sim_nca$id), time = dose_times) |>
    dplyr::mutate(amt = 4, treatment = "4 mg q6h")
)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "ug/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "mg")

intervals <- data.frame(
  start   = 0,
  end     = 24,
  cmax    = TRUE,
  tmax    = TRUE,
  auclast = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)
summary(nca_res)
```

# Comparison against the published NCA

The reference values are the paper's own NCA of the same virtual cohort:

* **Single dose** -- Table 4, which reports the literature reference
  ("Observed") and the GastroPlus PBPK prediction ("Predicted") for a single
  4 mg dose. The PBPK "Predicted" row is the more relevant comparator, because
  the PBPK model is what generated the dataset the popPK model was fitted to.
* **4 mg q6h** -- Supplementary Table S3, the arithmetic mean across its four
  virtual-population groups (Cmax 0.0134 / 0.0151 / 0.0139 / 0.0125 ug/mL,
  Tmax 19.2 / 19.0 / 18.8 / 19.0 h, AUC 0.212 / 0.222 / 0.177 / 0.199).

Table 4 and Table S3 both label their AUC column `ng.h/mL`, but the values are
in `ug*h/mL`: Table S3's AUCs are almost exactly four times Table 4's
single-dose AUC of 0.0553, as four doses of the same drug should be, and they
are three orders of magnitude away from being consistent with the Cmax column
in the same row. The unit label is an erratum (item 8); the values are used as
`ug*h/mL`. The published AUCs are extrapolated to infinity whereas the
simulated value stops at 24 h, but that difference is a few percent against
discrepancies of five- to six-fold, so it changes nothing below.

```{r nca-compare}
published <- tibble::tibble(
  treatment = c("Single 4 mg dose", "4 mg q6h"),
  cmax    = c(0.00919, mean(c(0.0134, 0.0151, 0.0139, 0.0125))),
  tmax    = c(1.28,    mean(c(19.2, 19.0, 18.8, 19.0))),
  auclast = c(0.0553,  mean(c(0.212, 0.222, 0.177, 0.199)))
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "treatment",
  units         = c(cmax = "ug/mL", tmax = "h", auclast = "ug*h/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated (Table 6 as printed) vs the paper's own published NCA (Table 4 Predicted row; Table S3 group mean). * marks rows differing by more than 20%."
)
attr(cmp, "footnote")
```

Every exposure row is starred: Cmax is low by a factor of 4.6 (single dose) to
6.4 (q6h), and AUC by a factor of 2.2 to 5.0.

The one row that passes -- q6h Tmax, +11% -- should not be read as agreement.
Both the published and the simulated Tmax fall late in a 24 h window simply
because both take their maximum after the last of four doses; the published
value is about 1 h after the 18 h dose while the simulated value is about 3.1 h
after it. The dimensionless peak-to-trough ratio is the shape statistic that
exposes this: about 3 in the published VPC against 1.29 simulated. A near-match
on an absolute Tmax can hide a complete failure of profile shape, and here it
does.

This is the central finding of this vignette, and it is not a simulation
artefact: the single-dose mass-balance gate above is exact to seven digits, so
the ODE system reproduces Table 6 faithfully. It is Table 6 that does not
reproduce the paper.

## Locating the discrepancy

The three parameters the paper itself flags or re-reports elsewhere are exactly
the three that move the profile. Table 6's Cl and V1 are the only two fixed
effects whose RSE fails the paper's own stated acceptance criterion, and Mtt is
re-reported in Table 7 on a completely different scale.

```{r locate}
t7_cl <- c(77.9, 69.9, 68.2)   # Table 7, Cl strata (L/h)
t7_v1 <- c(226, 276, 287)      # Table 7, V1 strata (L)
gm <- function(x) exp(mean(log(x)))

s3_auc <- c(0.212, 0.222, 0.177, 0.199)

tibble::tibble(
  Parameter = c("Mtt (h)", "Cl/F (L/h)", "V1/F (L)"),
  `Table 6 (encoded)` = c(9.50, 140, 77.5),
  `Table 7 EBE geometric mean` = c(
    sprintf("0.21 (F) / 0.30 (M)"),
    sprintf("%.1f", gm(t7_cl)),
    sprintf("%.1f", gm(t7_v1))
  ),
  `Ratio` = c(
    sprintf("%.0fx to %.0fx high", 9.50 / 0.30, 9.50 / 0.21),
    sprintf("%.2fx high", 140 / gm(t7_cl)),
    sprintf("%.2fx low", gm(t7_v1) / 77.5)
  )
) |>
  knitr::kable(caption = "Table 6 against Table 7, the paper's own empirical-Bayes geometric means for the same parameters.")
```

Two independent routes recover a clearance near Table 7's value rather than
Table 6's, both using only numbers the paper printed:

```{r cl-crosscheck}
tibble::tibble(
  Route = c(
    "Table 7 EBE geometric mean of the three Cl strata",
    "Total dose / AUC, from Table S3 (16 mg / mean AUC)",
    "Single dose / AUC, from Table 4 PBPK Predicted row (4 mg / 0.0553)",
    "Single dose / AUC, from Table 4 Observed row (4 mg / 0.0318)",
    "Table 6 as printed"
  ),
  `Implied Cl/F (L/h)` = c(
    gm(t7_cl), 16 / mean(s3_auc), 4 / 0.0553, 4 / 0.0318, 140
  )
) |>
  knitr::kable(digits = 1, caption = "Clearance implied by three independent published summaries, against Table 6's estimate.")
```

Three of the four independent routes cluster between 72 and 90 L/h; Table 6's
140 L/h is the outlier. The one route that agrees with Table 6 is the *literature
reference* AUC in Table 4's "Observed" row (126 L/h) -- not the PBPK prediction
that actually generated the fitted dataset.

# What this model is not

**It is not a corrected refit.** A model built from Table 6's ka, Q and V2 plus
Table 7's Cl and V1, and reading Mtt/Ktr on a minute rather than hour scale
(9.50 min = 0.158 h, Ktr = 9.00 /h, which leaves `n = 0.425` unchanged), lands
inside the published Table S3 range on Cmax, Tmax and AUC, and matches the
Figure 5 VPC on both peak timing and the dimensionless peak-to-trough ratio.
That reconciliation is reported here because it localises the errata, but it is
**not** what this package ships. Shipping it would mean publishing parameter
values under an author's name that the author never printed, chosen by us
because they fit better. The packaged entry is the published table; the errata
are documented; correspondence with the authors is the route to a corrected
version, and if one is published this entry should be revised to match it.

**It is not a covariate model.** The paper's title, abstract and conclusions
present a covariate popPK model, and Section 3.3.2 names five
covariate-parameter relationships. None of them has a published coefficient,
functional form, reference value or centering constant anywhere in the article
or its supplement, so none can be encoded. All eight screened covariates are
recorded in `covariatesDataExcluded` with their Table S5 statistics.

**It is not fitted to clinical data.** The 838 "observations" are GastroPlus
PBPK output. Any claim this model makes about real patients is inherited from
the PBPK model's assumptions, not tested against measurements.

**It is not three-compartment**, despite the model-selection statistics quoted
in the text belonging to the three-compartment row of Table S4 (item 4 below).

# Errata

Discrepancies found in the source publication. Items 1-3 are the ones that
change the simulated profile; items 4-8 are reporting defects; items 9-10 are
consequences of the published parameterisation and of the rxode2
implementation respectively.

1. **Mtt scale (Table 6 vs Table 7 and the Discussion).** Table 6 gives a
   population Mtt of 9.50 h. Table 7 gives empirical-Bayes geometric means of
   0.21 h (female) and 0.30 h (male) for the same parameter, and the Discussion
   discusses them as hours ("the geometric mean for females was 0.21 h, while
   for males it was 0.30 h"). Table 6 is 32-45x above its own individual
   estimates, which an IIV of 0.43 cannot bridge. Reading Table 6's Mtt/Ktr
   pair on a minute scale (9.50 min, 0.15 /min) reconciles them and leaves the
   derived chain length `n = 0.425` unchanged; a transposition of the two rows
   (Mtt 0.15 h, Ktr 9.50 /h) does so equally well and is indistinguishable in
   simulation. The paper as printed does not allow a choice between them.

2. **Cl/F (Table 6 vs Tables 4, 7 and S3).** Table 6's 140 L/h carries an RSE
   of 223%, which the paper itself flags as failing its stated < 30% criterion.
   Table 7's three strata give 77.9 / 69.9 / 68.2 L/h (geometric mean 71.9),
   `dose / AUC` on Table S3 gives 72-90 L/h, and `dose / AUC` on Table 4's PBPK
   prediction gives 72.3 L/h. Table 6 is about 1.9x high.

3. **V1/F (Table 6 vs Table 7).** Table 6's 77.5 L carries an RSE of 97.2%,
   also flagged by the authors. Table 7's three strata give 226 / 276 / 287 L
   (geometric mean 261.6), about 3.4x higher.

4. **Model-selection statistics attributed to the wrong model.** Section 3.3.1
   reports "BICc = -11,936.81 and -2LL = -12,022.77" for the two-compartment
   model with transit compartments. Supplementary Table S4 attributes exactly
   those two numbers to PK_S_08, the **three**-compartment model with transit
   compartments; the two-compartment row (PK_S_07) is BICc -9,980.17 and
   -2LL -10,045.39. The model is encoded as two-compartment because Table 6
   lists exactly the two-compartment parameter set (no Q2 or V3) and because
   the abstract, the Discussion and the Figure 5 caption all say
   two-compartment.

5. **Covariate effects reported without coefficients, and contradicted by the
   supplement.** Section 3.3.2 reports gender effects on Mtt and Cl, weight and
   BSA effects on V1 and Cl, and a health-status effect on Cl. No coefficient
   is published for any of them. Supplementary Table S5 supports the gender
   effects (p = 0.043 on Mtt, p = 0.034 on Cl) and the weight and BSA effects
   on V1 (both p < 0.001), but **contradicts** the weight and BSA effects on Cl
   (p = 0.724 and p = 0.646) and the health-status effect on Cl (p = 0.615).
   Separately, the Figure S4 caption states "Race has a significant impact on
   kel", which contradicts both Table S5 and the Discussion's explicit
   statement that race was not retained.

6. **IIV distribution stated as normal, evidenced as log-normal.** Section
   3.3.1 states "A normal distribution was used to describe IIV". Table 7's SD
   column for the same parameters is dimensionless and greater than 1
   throughout (1.37-3.89), i.e. those are geometric SDs, and they align with
   exponentiated Table 6 omegas (exp(0.32) = 1.38 against Mtt male SD 1.37;
   exp(0.36) = 1.43 against V1 SDs 1.54-1.84; exp(0.38) = 1.46 against Cl SDs
   1.45-2.28). The omegas are treated here as log-scale SDs and squared into
   `ini()` variances, which is also the Monolix default for positive PK
   parameters.

7. **Dosing interval stated two ways.** Section 2.2 and the Table S3 caption
   say 4 mg every 6 h. Section 3.1 says the regimen was "adjusted to 4 mg every
   4 h". Table S3's Tmax of 18.8-19.2 h (about 1 h after a dose at 18 h) and
   the Figure 5 VPC's 0-6 h x-axis both pin the interval at 6 h, so q6h is used.

8. **AUC units mislabelled.** Table 4 and Table S3 both label AUCinf as
   `ng.h/mL`, but the values are `ug*h/mL`. Table S3's values are four times
   Table 4's single-dose value, consistent with four doses, and are otherwise
   three orders of magnitude inconsistent with the Cmax column beside them.

9. **Negative implied transit-chain length for about a quarter of the
   population.** Table 6 places IIV on both Mtt and Ktr, so the derived chain
   length `n = Mtt * Ktr - 1` varies between subjects and is negative whenever
   `Mtt * Ktr < 1`. With the published omegas that is roughly 23% of the
   population (measured above). A negative number of transit compartments has
   no physical interpretation; rxode2 evaluates the gamma-form input anyway
   (only `n > -1` is required), but a small number of the most extreme draws
   are numerically stiff enough to fail to solve and are dropped from the NCA
   with the count reported.

10. **Closed-form transit under q6h dosing (implementation, not source).**
    rxode2's `transit()` is driven by time-since and amount-of the most recent
    dose, so each dose restarts the transit input and discards the un-absorbed
    remainder of the previous one. With Mtt (9.50 h) longer than the dosing
    interval (6 h) that discards a large fraction of the administered dose, as
    quantified above. Single-dose mass balance is exact, so this affects only
    multiple-dose simulations. A user who needs exact multiple-dose
    superposition should replace `transit()` with an explicit chain of transit
    ODEs.

11. **Correction notice.** The PDF carries a publisher's "Correction Statement"
    noting the article "has been republished with a minor change" that "does
    not affect the scientific content". No parameter-revising erratum was found
    on the journal landing page, in PubMed, or in the supplementary material at
    the time of extraction.

# Assumptions and deviations

* **Table 6 is encoded verbatim, including the three values its own tables
  contradict.** No parameter was tuned, substituted or reconciled. The
  alternative reconciliation is described above and deliberately not shipped.
* **Log-normal IIV** assumed from the Table 7 geometric SDs rather than the
  stated normal distribution (erratum 6). All seven omegas are squared into
  variances and left diagonal, per Section 3.3.2's statement that no
  significant parameter correlations were found.
* **`combined1()` residual error**, i.e. additive and proportional SDs adding
  linearly, from the Section 3.3.1 error equation. nlmixr2's default for
  `add() + prop()` adds them in quadrature, which would not match the paper.
* **Bioavailability is not estimated**, so Cl, V1, Q and V2 are apparent. The
  PBPK cohort's own F was 39.6-46.0% (Table S3), but that is a property of the
  data-generating model, not a popPK estimate, and is not encoded.
* **The transit chain length `n = 0.425` is derived**, not printed. It follows
  from Monolix's `n = Mtt * Ktr - 1` parameterisation and reproduces both
  printed values exactly, as verified above.
* **q6h chosen over "every 4 h"** on the evidence of Table S3's Tmax and the
  Figure 5 VPC axis (erratum 7).
* **Two compartments chosen over three** on the evidence of Table 6's parameter
  set and the abstract, Discussion and figure caption (erratum 4).
* **No covariate effects encoded** because none has a published coefficient
  (erratum 5). All eight screened covariates are preserved in
  `covariatesDataExcluded`. The three-level health-status covariate is recorded
  under `HEPIMP_MILD` (Child-Pugh A is a registered alias); its obese level has
  no registered canonical at the correct severity and is documented in prose
  rather than under an invented column name.
* **A plain-ODE twin of the packaged model is used for all simulation**, because
  `transit()` evaluates to zero when the model is solved through the `rxUi`
  object on the installed rxode2 version. The workaround is asserted, not
  assumed, in the "Model handle" section, and the assertion will fail once
  rxode2 fixes the underlying defect.
* **Race is documented as two indicators** (`RACE_ASIAN`, `RACE_CHINESE`)
  against an American reference, because the virtual cohort has three race
  levels. The source screened race as a single categorical.
