---
title: "Berotralstat (Mathis 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Berotralstat (Mathis 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Berotralstat (BCX7353, Orladeyo) is an oral, once-daily small-molecule
inhibitor of plasma kallikrein, approved for prophylaxis of attacks of
hereditary angioedema (HAE) in adults and adolescents aged 12 years and older.
Mathis 2022 pooled 10,437 plasma concentrations from 771 healthy subjects and
HAE patients in 13 phase 1-3 studies and described them with a
three-compartment model with first-order absorption, an absorption lag time and
linear elimination. Body weight acts on clearance and central volume, and
relative bioavailability rises with dose.

```{r model}
mod <- rxode2::rxode(readModelDb("Mathis_2022_berotralstat"))
```

* Citation: `r mod$reference`
* Article: <https://doi.org/10.1111/cts.13233>
* Supporting Information (EuropePMC, open access):
  <https://europepmc.org/article/PMC/PMC9010267>

Table 1 of the paper gives the final fixed effects. Supporting Information
Text S2 is the NONMEM control stream of the final model, which fixes the
structure: `ADVAN12`, the `$PK` covariate equations and a combined
additive-plus-proportional `$ERROR` block. The random-effect variances are a
special case, described under *Assumptions and deviations* below.

# Population

```{r population}
str(mod$population)
```

The 771 subjects (Results, Data summary; Supporting Information Table S3) were
aged 12-74 years (mean 38.1, SD 13.9) and weighed 40.1-150 kg (median 75.1 kg).
53.7 percent were male. 83.3 percent were White, 8.7 percent Asian, 4.5 percent
Black or African American and 2.7 percent other races. About half were HAE
patients: 400 of 771 came from the four patient studies BCX7353-203, -204, -301
and -302. The rest were healthy subjects, including small renal- and
hepatic-impairment cohorts. Only 16 subjects were 17 years old or younger.
Doses ranged from single doses of 10-900 mg to once-daily doses of 110-450 mg.

# Source trace

| Model element | Value | Source |
|---|---|---|
| Structure: 3 compartments, first-order absorption, lag time, linear elimination | -- | Results, Final model; Figure 1; Text S2 `$SUBROUTINE ADVAN12` |
| `lcl` CL | 47.3 L/h | Table 1 |
| `lvc` V2 | 1650 L | Table 1 |
| `lk12` K23 | 0.0812 1/h | Table 1 |
| `lk21` K32 | 0.0309 1/h | Table 1 |
| `lk13` K24 | 0.00281 1/h | Table 1 |
| `lk31` K42 | 0.00136 1/h | Table 1 |
| `lka` Ka | 1.12 1/h | Table 1 |
| `ltlag` ALAG1 | 0.468 h | Table 1 |
| `e_wt_cl` | 0.480 | Table 1; Text S2 `TVCL= THETA(1)*CWTKG**THETA(15)` |
| `e_wt_vc` | 1.00 | Table 1; Text S2 `TVV2=THETA(2)*CWTKG**THETA(13)` |
| Weight reference | 75.2 kg | Table 1 footnote; Text S2 `CWTKG =WTKG/75.2` |
| `e_dose_f`, F1 = (Dose/300)^THETA(11) | 0.497 | Table 1; Equation 1; Text S2 `F1=1*(DOSE/CDOSE)**THETA(11)`, `CDOSE = 300` |
| `etalcl`, `etalvc`, `etalka`, `etaltlag` block | 0.141588 ... 0.245715 | Text S2 `$OMEGA BLOCK(4)` (initial estimates) |
| `eta_study_lcl` | 0.0817099 | Text S2 `$OMEGA` ETA(5), `$LEVEL STDY=(5[1])` (initial estimate) |
| `addSd` | 0.483 ng/mL | Table 1; Text S2 `W=SQRT(ADDERR**2+PROPERR**2*IPRE**2)`, `$SIGMA 1 FIX` |
| `propSd` | 0.286 | Table 1 |
| `Cc = 1000 * central / vc` | -- | Text S2 `S2 = V2/1000` (dose mg, concentration ng/mL) |

# Typical-value checks

## Terminal half-life

The paper reports a terminal half-life of 566 h from the final model (Results,
Final model and model evaluation). That is a property of the typical
micro-constants alone, so it is checked from the eigenvalues of the
disposition rate matrix.

```{r halflife}
p <- exp(unlist(mod$theta[c("lcl", "lvc", "lk12", "lk21", "lk13", "lk31")]))
kel <- p[["lcl"]] / p[["lvc"]]
rate_matrix <- matrix(
  c(-(kel + p[["lk12"]] + p[["lk13"]]), p[["lk21"]], p[["lk31"]],
    p[["lk12"]], -p[["lk21"]], 0,
    p[["lk13"]], 0, -p[["lk31"]]),
  nrow = 3, byrow = TRUE
)
half_lives <- sort(log(2) / -eigen(rate_matrix)$values)
half_lives

# Deterministic: depends only on the printed Table 1 values. The residual
# 0.5 percent is the rounding of K24 and K42 to three significant figures.
stopifnot(abs(half_lives[3] / 566 - 1) < 0.01)
```

The three phases have half-lives of about 5 h, 96 h and 569 h. The middle one
matches the 93-h NCA half-life of the single-dose studies. The paper's
Discussion explains why: those studies were too short to see the third phase.

## Steady-state AUC against its closed form

At steady state the dosing-interval AUC of a linear model is
F x Dose / CL. This checks the solve, the bioavailability term and the
mg-to-ng/mL scaling together.

```{r closed-form}
typ <- rxode2::zeroRe(mod)
obs_times <- seq(0, 24, by = 0.25)

make_events <- function(wt, dose = 150, ss = 0) {
  rbind(
    data.frame(time = 0, evid = 1L, amt = dose, cmt = "depot",
               ss = ss, ii = if (ss == 1) 24 else 0),
    data.frame(time = obs_times, evid = 0L, amt = 0, cmt = "central",
               ss = 0, ii = 0)
  ) |>
    dplyr::mutate(WT = wt, DOSE_BEROTRALSTAT_MG = dose)
}

trap_auc <- function(time, conc) {
  sum(diff(time) * (head(conc, -1) + tail(conc, -1)) / 2)
}

ss_typ <- rxode2::rxSolve(typ, make_events(75.2, ss = 1), returnType = "data.frame")
auc_ss_solve <- trap_auc(ss_typ$time, ss_typ$Cc)
auc_ss_closed <- 1000 * (150 / 300)^0.497 * 150 / 47.3
c(solve = auc_ss_solve, closed_form = auc_ss_closed)

# Same drawn (typical) parameters on both sides: the gap is trapezoidal error on
# a 0.25 h grid, so the bound is tight.
stopifnot(abs(auc_ss_solve / auc_ss_closed - 1) < 0.005)
```

# Virtual cohort

Table 2 of the paper simulates 1000 HAE patients per scenario at the commercial
dose of 150 mg once daily. Its adult scenarios are four weight bands, with
weights drawn from a log-normal distribution whose mean and variance come from
the analysis dataset. Here each band gets 200 subjects. Weights are drawn from a
log-normal with the Table S3 all-subject mean 77.7 kg and SD 17.2 kg, truncated
to the band.

The model has no age effect, so the paper's adolescent scenario differs from the
adults only in its weight distribution. That scenario uses CDC weight-for-age
charts, whose parameters the paper does not print, so it is not re-simulated.
The adolescent row is checked separately below through the weight it implies.

```{r cohort}
n_per_band <- 200
bands <- tibble::tribble(
  ~band,                               ~lo, ~hi,
  "Underweight adults (40-60 kg)",      40,  60,
  "Low weight adults (60-80 kg)",       60,  80,
  "Normal weight adults (80-100 kg)",   80, 100,
  "Overweight adults (100-120 kg)",    100, 120
)

sdlog <- sqrt(log(1 + (17.2 / 77.7)^2))
meanlog <- log(77.7) - sdlog^2 / 2

rxode2::rxSetSeed(20220401)
cohort <- bands |>
  dplyr::rowwise() |>
  dplyr::reframe(
    band = band,
    WT = stats::qlnorm(
      stats::runif(n_per_band,
                   stats::plnorm(lo, meanlog, sdlog),
                   stats::plnorm(hi, meanlog, sdlog)),
      meanlog, sdlog
    )
  ) |>
  dplyr::mutate(id = dplyr::row_number())

cohort |>
  dplyr::group_by(band = factor(band, levels = bands$band)) |>
  dplyr::summarise(n = dplyr::n(), median_wt = stats::median(WT),
                   min_wt = min(WT), max_wt = max(WT), .groups = "drop") |>
  dplyr::rename("Weight band" = band, "N" = n, "Median weight (kg)" = median_wt,
                "Min (kg)" = min_wt, "Max (kg)" = max_wt) |>
  knitr::kable(digits = 1)
```

# Simulation

Each subject is simulated twice: after the first 150 mg dose, and over one
dosing interval at true steady state (`ss = 1`). The steady-state option
matters here. With a 569-h terminal half-life, even several months of daily
dosing falls short of steady state.

Table 2's geometric CVs for AUC at steady state are 36.7-38.6 percent. Within a
weight band, the CV of that AUC reflects only CL, through F x Dose / CL. The
between-subject variance on CL alone (0.1416, a CV of 39 percent) accounts for
it. Adding the between-study variance (0.0817) would push it to about 50
percent. The Table 2 replication therefore sets the between-study eta to zero,
which appears to be what the paper's own simulations did. Concentrations are
individual predictions without residual error.

```{r simulate}
mod_no_study <- mod |> rxode2::ini(eta_study_lcl = fix(0))

build_events <- function(ss) {
  cohort |>
    dplyr::rowwise() |>
    dplyr::reframe(id = id, band = band, make_events(WT, ss = ss)) |>
    as.data.frame()
}

ev_first <- build_events(ss = 0)
ev_ss <- build_events(ss = 1)

rxode2::rxSetSeed(20220402)
sim_first <- rxode2::rxSolve(mod_no_study, ev_first, keep = "band",
                             returnType = "data.frame") |>
  dplyr::mutate(regimen = "First dose")
# The steady-state solve iterates to convergence over a 569-h terminal phase.
# With the default step limit a few subjects with extreme absorption etas stop
# with "excess work done" and come back as NA, which PKNCA would then drop
# silently. A larger step limit solves them all; the gate below fails loudly
# if any subject is still missing.
rxode2::rxSetSeed(20220402)
sim_ss <- rxode2::rxSolve(mod_no_study, ev_ss, keep = "band",
                          maxsteps = 500000L, returnType = "data.frame") |>
  dplyr::mutate(regimen = "Steady state")
sim <- dplyr::bind_rows(sim_first, sim_ss)
stopifnot(!anyNA(sim$Cc), dplyr::n_distinct(sim_ss$id) == nrow(cohort))
```

## Replicates Figure 4

```{r figure-4, fig.height = 4}
sim |>
  dplyr::group_by(regimen, band, time) |>
  dplyr::summarise(median_cc = stats::median(Cc), .groups = "drop") |>
  ggplot(aes(time, median_cc, colour = factor(band, levels = bands$band))) +
  geom_line() +
  facet_wrap(~regimen, scales = "free_y") +
  labs(x = "Time after dose (h)", y = "Median berotralstat (ng/mL)",
       colour = NULL,
       caption = "Replicates Figure 4 of Mathis 2022 (150 mg once daily).") +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(ncol = 2))
```

# PKNCA validation

```{r pknca}
run_nca <- function(dat) {
  conc <- dat |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::select(id, time, Cc, band)
  # One time-zero row per subject. After the first dose it is 0; at steady state
  # it is already in the solve output (the pre-dose trough), and distinct()
  # keeps that row.
  conc <- dplyr::bind_rows(
    conc,
    conc |> dplyr::distinct(id, band) |> dplyr::mutate(time = 0, Cc = 0)
  ) |>
    dplyr::distinct(id, band, time, .keep_all = TRUE) |>
    dplyr::arrange(id, time)
  doses <- conc |> dplyr::distinct(id, band) |> dplyr::mutate(time = 0, amt = 150)
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(conc, Cc ~ time | band + id),
    PKNCA::PKNCAdose(doses, amt ~ time | band + id),
    intervals = data.frame(start = 0, end = 24, cmax = TRUE, auclast = TRUE)
  ))
  as.data.frame(res)
}

# One PKNCA call per band and regimen keeps each call to 200 subjects.
nca <- sim |>
  dplyr::group_split(regimen, band) |>
  lapply(function(d) dplyr::mutate(run_nca(d), regimen = d$regimen[1])) |>
  dplyr::bind_rows()

nca_wide <- nca |>
  dplyr::filter(PPTESTCD %in% c("cmax", "auclast")) |>
  dplyr::select(band, id, regimen, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
```

## Comparison against Table 2

```{r table-2}
gcv <- function(x) 100 * sqrt(exp(stats::var(log(x))) - 1)

published <- tibble::tribble(
  ~band,                              ~auc_first, ~cmax_first, ~auc_ss, ~cmax_ss, ~cv_auc_ss, ~rac,
  "Underweight adults (40-60 kg)",           816,          67,    2574,      155,       36.7, 3.15,
  "Low weight adults (60-80 kg)",            644,          51,    2211,      129,       38.3, 3.43,
  "Normal weight adults (80-100 kg)",        509,          40,    1951,      111,       37.0, 3.83,
  "Overweight adults (100-120 kg)",          438,          34,    1818,      101,       36.9, 4.15
)

simulated <- nca_wide |>
  dplyr::group_by(band) |>
  dplyr::summarise(
    auc_first = exp(mean(log(auclast[regimen == "First dose"]))),
    cmax_first = exp(mean(log(cmax[regimen == "First dose"]))),
    auc_ss = exp(mean(log(auclast[regimen == "Steady state"]))),
    cmax_ss = exp(mean(log(cmax[regimen == "Steady state"]))),
    cv_auc_ss = gcv(auclast[regimen == "Steady state"]),
    .groups = "drop"
  ) |>
  dplyr::mutate(rac = auc_ss / auc_first)

cmp <- dplyr::inner_join(simulated, published, by = "band",
                         suffix = c("_sim", "_pub")) |>
  tidyr::pivot_longer(-band, names_to = c("metric", ".value"),
                      names_pattern = "(.*)_(sim|pub)") |>
  dplyr::mutate(pct_diff = 100 * (sim - pub) / pub)

metric_labels <- c(
  auc_first = "AUC0-24, first dose (ng*h/mL)",
  cmax_first = "Cmax, first dose (ng/mL)",
  auc_ss = "AUCtau, steady state (ng*h/mL)",
  cmax_ss = "Cmax, steady state (ng/mL)",
  cv_auc_ss = "Geometric CV of AUCtau,ss (%)",
  rac = "Accumulation ratio"
)

cmp |>
  dplyr::arrange(match(band, bands$band), match(metric, names(metric_labels))) |>
  dplyr::mutate(metric = metric_labels[metric]) |>
  dplyr::rename("Weight band" = band, "NCA parameter" = metric,
                "Simulated" = sim, "Mathis 2022 Table 2" = pub,
                "% difference" = pct_diff) |>
  knitr::kable(digits = 2, caption = paste(
    "Simulated geometric means (200 subjects per band) against the paper's",
    "Table 2 (1000 per band), 150 mg once daily."
  ))
```

```{r table-2-gate}
exposure <- cmp |> dplyr::filter(metric %in% c("auc_first", "cmax_first", "auc_ss", "cmax_ss"))
summary(exposure$pct_diff)

# Assert on the centre and on a robust quantile, not on the extremes: the
# cohort is a random draw and its tails are not reproducible across rxode2
# builds. A mis-transcribed CL, V2, bioavailability exponent or unit moves the
# whole set by tens of percent.
stopifnot(
  abs(stats::median(exposure$pct_diff)) < 7,
  stats::quantile(abs(exposure$pct_diff), 0.9) < 15
)

# The accumulation ratio is a level-independent ratio of two AUCs from the same
# subjects. It tests the distribution rate constants and the steady-state
# solve, independently of CL and F.
rac <- cmp |> dplyr::filter(metric == "rac")
stopifnot(abs(stats::median(rac$pct_diff)) < 7)

# Variability: the CV of AUCtau,ss is driven by the CL variance. With n = 200
# the sampling error of a CV near 38 percent is about 2 points.
cv <- cmp |> dplyr::filter(metric == "cv_auc_ss")
stopifnot(abs(stats::median(cv$sim - cv$pub)) < 6)
```

All sixteen geometric means fall within 7 percent of Table 2, with a
median difference near zero. The accumulation ratio rises with weight, as in
the paper (Table 2: 3.15 to 4.15). The simulated ratios run 1-6 percent above
the published ones in every band, because the simulated steady-state AUCs are
slightly higher. A likely cause is that the paper's "steady state" was a long
but finite run of daily doses rather than the true steady state used here; with
a 569-h terminal phase the two differ by a few percent. The geometric CV of AUC
at steady state matches Table 2 within the sampling error of 200 subjects
(median gap about 1 point). The control-stream variance on CL, an initial
estimate, therefore reproduces the published spread.

## Adolescents: the weight Table 2 implies

Within the model, AUC at steady state scales as WT^-0.480 and nothing else
differs between adolescents and adults. Inverting the low-weight-adult row gives
the typical weight that each Table 2 row implies.

```{r adolescent}
table2_auc_ss <- c(
  "Underweight adults (40-60 kg)" = 2574,
  "Low weight adults (60-80 kg)" = 2211,
  "Normal weight adults (80-100 kg)" = 1951,
  "Overweight adults (100-120 kg)" = 1818,
  "Adolescents (12-18 years)" = 2515
)
implied_wt <- 75.2 * (table2_auc_ss / auc_ss_closed)^(-1 / 0.480)
round(implied_wt, 1)
```

The inversion is sensitive: the exponent 1/0.480 turns a 1 percent AUC
difference into a 2 percent weight difference, so the adult rows land near the
centre or upper edge of their bands rather than exactly at the typical weight.
The adolescent row implies about 60 kg, just above the underweight adults (about
57 kg). This matches the paper's statement that adolescents and underweight
adults have almost equal exposures, and in this model weight is the only
covariate that could produce it.

# Stochastic VPC with all variability

For completeness, the full model is simulated as shipped: both the between-subject
and the between-study random effects, plus residual error. The cohort is 200
subjects at 75.2 kg, after the first 150 mg dose.

```{r vpc}
vpc_events <- do.call(rbind, lapply(seq_len(200), function(i) {
  dplyr::mutate(make_events(75.2), id = i)
}))
rxode2::rxSetSeed(20220403)
vpc <- rxode2::rxSolve(mod, vpc_events, returnType = "data.frame")

vpc |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    p05 = stats::quantile(sim, 0.05), p50 = stats::median(sim),
    p95 = stats::quantile(sim, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(time, p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.25) +
  geom_line() +
  labs(x = "Time after first 150 mg dose (h)", y = "Berotralstat (ng/mL)",
       caption = "Median and 90% prediction interval, 75.2 kg, full random effects.") +
  theme_bw()
```

# Assumptions and deviations

* **Random-effect variances are control-stream initial estimates.** The paper
  states which parameters carry between-subject variability (CL, V2, Ka and the
  lag time, in one correlated block) and that a between-study random effect acts
  on CL. It does not print the final variances, in Table 1 or in the
  Supporting Information. The only values published are the `$OMEGA` initial
  estimates in the Text S2 final-model control stream, and the model uses those
  verbatim. They appear to be near-converged: the `$THETA` initials in the same
  file are within 5 percent of the Table 1 finals for every parameter except
  the poorly determined K24 and K42 (11 and 18 percent; for example CL 47.0 vs
  47.3 L/h, V2 1603 vs 1650 L), and the AUC CVs simulated here sit a few points
  above Table 2. The implied CVs are about 39 percent on CL, 34 percent
  on V2, 101 percent on Ka and 53 percent on the lag time, plus 29 percent
  between studies on CL.
* **Between-study variability is drawn per subject.** In NONMEM ETA(5) is shared
  by every subject in a study (`$LEVEL STDY=(5[1])`). In a plain `rxSolve` it is
  drawn per subject, which adds to the between-subject spread on CL. Set
  `eta_study_lcl` to zero to reproduce the paper's Table 2, as above. To mimic a
  multi-study trial, draw it once per study and pass it in as a data column.
* **Dose covariate.** `DOSE_BEROTRALSTAT_MG` must equal the dose amount. In the
  source, API-in-capsule doses are expressed as the HCl salt and
  blend-in-capsule (commercial) doses as the free base. The two were
  bioequivalent at those nominal doses (Table S2).
* **Figure 1 caption.** The caption says "higher doses predicting lower
  bioavailability". Equation 1, the positive THETA(11) of 0.497, the Study
  Highlights and the Discussion all say the opposite: bioavailability rises with
  dose. The model follows the equation.
* **Dose range.** The Methods quote single doses of 10-900 mg. Supporting
  Information Table S1 lists study BCX7353-101 Part 1 doses up to 1000 mg. The
  population metadata uses the Methods range.
* **BLQ handling and the assay switch.** The control stream fits BLQ records
  with the M3 likelihood. It also computes an `ASSAY` flag that the final
  `$ERROR` block never uses. Neither affects simulation.
