---
title: "Nilotinib tablets vs capsules (Mauro 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nilotinib tablets vs capsules (Mauro 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Mauro 2025 reports two sequential fits of the same structural model. Both are
distributed, because each drove a different set of published simulations.

```{r models}
uiFinal <- rxode2::rxode(readModelDb("Mauro_2025_nilotinib"))
uiLowfat <- rxode2::rxode(readModelDb("Mauro_2025_nilotinib_lowfat"))
```

* Citation: `r uiFinal$reference`
* Article: <https://doi.org/10.1007/s00280-025-04777-6>
* Supplement: <https://doi.org/10.1007/s00280-025-04777-6> (Supplementary
  Information, `280_2025_4777_MOESM1_ESM.docx`; carries the model-development
  narrative, the F1 equations, and the two parameter tables)

| Model file | Source table | Studies | Prandial states covered | Published result it drove |
|---|---|---|---|---|
| `Mauro_2025_nilotinib` | Suppl. Table 2 ("Final Model") | 13 | fasted; high-fat meal 0.5 h pre-dose; high-fat meal 2 h pre-dose | Table 1 fasted steady-state bioequivalence |
| `Mauro_2025_nilotinib_lowfat` | Suppl. Table 3 ("Updated Model") | 14 | the three above, plus low-fat meal 2 h pre-dose | Table 2 low-fat-meal steady-state exposures |

The Updated Model adds study 052-23 (190 mg tablets vs 400 mg capsules under
modified fasting with a low-fat meal) and keeps the structure of the Final Model,
so its shared parameters shift only slightly (CL 34.5 -> 34.3 L/h, V1 557 ->
551 L). Per `references/replicate-author-structure.md` both are extracted, one
file each, because both are final models for their own published purpose; the
paper's fasted bioequivalence conclusion rests on the first and its low-fat-meal
food-effect conclusion on the second.

# Population

The pooled analysis dataset comprises 30,089 nilotinib plasma concentrations
from 502 healthy adult subjects across 14 single-dose, relative-bioavailability
and food-effect crossover studies (Suppl. Table 1), with 1,311
below-limit-of-quantification records excluded. Nilotinib tablets were given at
50 to 300 mg and capsules at 50 to 400 mg, always as a single dose, with each
dose delivered as two units (for example 190 mg = 2 x 95 mg tablets, 400 mg =
2 x 200 mg capsules). Only 39 participants were female, contributed by studies
006-23, 007-23 and 128-22; every other study enrolled healthy men only, which is
why sex effects are identifiable only through those three studies. The trials
were run in Phase 1 units in Hyderabad, India.

Baseline demographic distributions (age, weight, laboratory values) are not
tabulated in the paper or in its Supplemental Information, so the virtual cohort
below fixes body weight at the assumed 70 kg allometric reference and splits sex
50/50 to match the paper's own simulation cohort ("A dataset of 50 healthy
subjects (50% male/50% female) was randomly selected from the analysis dataset").

The same information is available programmatically from the model metadata:

```{r population}
str(uiFinal$population)
```

# Model structure

A two-compartment model with linear elimination. Absorption has two parallel
routes, both fed by the same dose amount:

1. **Primary**, a zero-order input directly into `central` of duration `D1` with
   lag `ALAG1` and relative bioavailability `F1`. One-, mixed zero-then-first-
   order and transit-chain alternatives were tested and rejected (Suppl.
   Information, "Initial Model Development").
2. **Secondary** (`depot2`), an empirical first-order route with its own lag,
   carrying a fraction `1 - 0.888 = 0.112` of `F1`. It was added to describe an
   increase in concentrations at a time since last dose of about 12 and 24 h
   (delta OFV -575.6). Total absorbed drug is therefore `1.112 x F1 x dose`.

In the event table each administration is **two dose records with the same
`amt`**: one into `central` with `rate = -2` (modelled duration) and one into
`depot2`. The model files declare `dosing <- c("central", "depot2")`.

Relative bioavailability follows the supplement's equations verbatim:

```
F1_capsule = 1 * (DOSE/400)^theta10
F1_tablet  = 1 * (1 + theta6) * (DOSE/200)^theta10   for DOSE <= 200 mg
F1_tablet  = 1 * (1 + theta6)                        for DOSE >  200 mg
```

implemented as `min(DOSE, 200)` in the tablet branch so tablet exposure stays
dose-proportional above 200 mg while capsule exposure remains less than dose
proportional.

## Encoding the four prandial states

The paper defines four mutually exclusive prandial states (Methods, "Prandial
state definitions"), and all four carry different bioavailability effects, so all
four must be separable. Meal *composition* and meal *timing* are orthogonal here:
the paper's "fed" and "modified fasting with a high-fat meal" states use the
**same** high-fat meal and differ only in whether the dose followed the start of
the meal by about 0.5 h or about 2 h. Composition indicators alone therefore
cannot tell them apart, which is why this extraction registers the new canonical
covariate `MEAL_PREDOSE_2H` (see `inst/references/covariate-columns.md`).

| Paper's prandial state | `FED` | `FED_HIGHFAT` | `FED_LOWFAT` | `MEAL_PREDOSE_2H` |
|---|---|---|---|---|
| fasted (overnight fast >= 10 h) | 0 | 0 | 0 | 0 |
| "fed" (high-fat meal started 0.5 h pre-dose) | 1 | 1 | 0 | 0 |
| "modified fasting" + high-fat meal (2 h pre-dose) | 1 | 1 | 0 | 1 |
| "modified fasting" + low-fat meal (2 h pre-dose) | 1 | 0 | 1 | 1 |

`MEAL_PREDOSE_2H` is the mirror image of the existing `MEAL_DELAY_1H` canonical:
`MEAL_DELAY_<n>` marks a meal taken *after* the dose, `MEAL_PREDOSE_<n>` a meal
taken *before* it. The bare phrase "modified fasting" is deliberately **not**
used to define the covariate, because it is not portable between papers -- Mauro
2025 uses it for *meal first, dose 2 h later*, while other nilotinib-era papers
use the same phrase for *dose first, food withheld for 2 h afterwards*.

# Source trace

Every `ini()` entry in both model files carries an in-file comment naming its
source row. The table below collects the Final Model (Suppl. Table 2); the
Updated Model (Suppl. Table 3) is the identically-structured table on the
following supplement page, plus four low-fat-meal rows.

| Equation / parameter | Value | Source location |
|---|---|---|
| 2-compartment, linear elimination, zero-order absorption with lag | n/a | Results, "Final population PK model"; Suppl. Info, "Initial Model Development" |
| Secondary depot with first-order absorption and lag | n/a | Suppl. Info, "Initial Model Development" (delta OFV -575.6) |
| `F1` equations (formulation, dose power, 200 mg tablet cap) | n/a | Suppl. Info, "Initial Model Development", displayed equations |
| `lcl` (CL) | 34.5 L/h | Suppl. Table 2, Typical Values |
| `lvc` (V1) | 557 L | Suppl. Table 2, Typical Values |
| `lq` (Q) | 4.54 L/h | Suppl. Table 2, Typical Values |
| `lvp` (V2) | 223 L | Suppl. Table 2, Typical Values |
| `ld1` (D1) | 2.32 h | Suppl. Table 2, Typical Values |
| `ltlag` (ALAG1) | 0.0988 h | Suppl. Table 2, Typical Values |
| `lka2` (secondary depot ka) | 0.140 1/h | Suppl. Table 2, Typical Values |
| `ltlag2` (secondary depot lag) | 5.35 h | Suppl. Table 2, Typical Values |
| `lfdepot2` (fraction of F1 re-absorbed) | 1 - 0.888 | Suppl. Table 2, Typical Values |
| `lfdepot` (capsule F1 anchor) | 1 (fixed) | Suppl. Info F1 equations |
| `e_form_tablet_fdepot` | 0.988 | Suppl. Table 2, Covariate Effects |
| `e_dose_fdepot` | -0.701 | Suppl. Table 2, Covariate Effects |
| `e_form_tablet_d1` | -0.423 | Suppl. Table 2, Covariate Effects |
| `e_wt_cl` | -0.727 | Suppl. Table 2, Covariate Effects |
| `e_hfmeal_cap_fdepot` | 1.81 | Suppl. Table 2, "Food (modified fasted) effect on F1 for nilotinib capsules" |
| `e_hfmeal05h_tab{50,142,190,240,300}_fdepot` | 0.166, 0.484, 0.617, 0.748, 0.614 | Suppl. Table 2, "Food (fed) effect on F1 at <dose> for nilotinib tablets" |
| `e_hfmeal2h_tab142_fdepot` | 0.585 | Suppl. Table 2, "Food (modified fasted) effect on F1 at 142 mg for nilotinib tablets" |
| `e_hfmeal2h_tab190_fdepot` | 0.603 | Suppl. Table 2, row labelled "... at 190 mg for nilotinib capsules" (label typo; see Errata) |
| `e_fed_tab_d1`, `e_fed_cap_d1` | 2.91, 0.746 | Suppl. Table 2, "Food effect on D1 ..." |
| `e_fed_tab_tlag`, `e_fed_cap_tlag` | 2.98, 7.79 | Suppl. Table 2, "Food effect on ALAG1 ..." |
| `e_fed_vc`, `e_fed_q` | -0.248, 2.07 | Suppl. Table 2, "Food effect on V1 / Q" |
| `e_sexf_cl` | -0.227 | Suppl. Table 2, "Sex (female) covariate effect on CL" |
| `e_sexf_fdepot` | 0.116 | Suppl. Table 2, "Sex (female*) covariate effect on F1" + footnote |
| `etalcl + etalvc` block | 0.108, 0.0628, 0.144 | Suppl. Table 2, Between Subject Variability |
| `etaiov_vc_*` (IOV on V1) | 0.157 | Suppl. Table 2, Inter-Occasion Variability |
| `propSd`, `addSd` | 0.230, 36.6 ng/mL | Suppl. Table 2, Residual Error |
| `e_lfmeal2h_tab_fdepot`, `e_lfmeal2h_cap_fdepot` | 0.294, 0.512 | Suppl. Table 3, low-fat-meal rows (+ Suppl. Info text "29.4% and 51.2% for nilotinib tablets and capsules respectively") |
| `e_lfmeal2h_tab_d1`, `e_lfmeal2h_cap_tlag` | 0.948, 3.25 | Suppl. Table 3, low-fat-meal rows (+ Suppl. Info text "an effect on D1 for nilotinib tablets and on lag-time for capsules") |

# Virtual cohort

The paper's steady-state simulations dosed twice daily for 10 days and computed
AUCss, Cmaxss and Cminss over the 0-12 h interval after the last dose, in 50
resampled subjects split 50/50 by sex. The cohort below follows that design with
100 subjects per arm across 14 arms.

Every pooled study was a **crossover** design, so the arms are simulated as a
crossover too: each arm is solved separately with the random-number seed reset
first, which gives subject `k` the same clearance, volume, occasion and
residual-error draws in every arm. That is both the faithful design and the
statistically sharper one -- the published quantities below are all *ratios*
between arms, and pairing removes the between-arm sampling noise that would
otherwise dominate them at this cohort size.

```{r cohort}
simSeed <- 20250422L

nPerArm <- 100L
tau <- 12          # BID dosing interval (h)
nDose <- 20L       # 10 days BID, as in Mauro 2025 Methods, "Simulations"
tLast <- tau * (nDose - 1L)
obsGrid <- seq(0, tau, by = 0.5)

# A realistic bioequivalence sampling schedule inside the last interval; used
# only for the residual-error Cmax diagnostic further down. Every element is on
# obsGrid, so no extra solve is needed.
beGrid <- c(0, 0.5, 1, 1.5, 2, 2.5, 3, 4, 6, 8, 10, 12)
stopifnot(all(beGrid %in% obsGrid))

# Build one arm's event table. Each administration is two dose records with the
# same amt: the lagged zero-order input into central (rate = -2 selects the
# modelled duration D1) and the secondary depot. Only numeric covariate columns
# go into the event table; arm labels are attached after the solve.
makeArm <- function(dose, tablet, fed, highfat, lowfat, predose2h, n = nPerArm) {
  subj <- tibble(
    id = seq_len(n),
    SEXF = as.integer(seq_len(n) %% 2L == 0L),
    WT = 70
  )
  doses <- subj |>
    tidyr::crossing(time = seq(0, tLast, by = tau),
                    cmt = c("central", "depot2")) |>
    mutate(amt = dose, evid = 1L, rate = if_else(cmt == "central", -2, 0))
  obs <- subj |>
    tidyr::crossing(time = tLast + obsGrid) |>
    mutate(cmt = "central", amt = 0, evid = 0L, rate = 0)
  bind_rows(doses, obs) |>
    mutate(DOSE = dose, FORM_TABLET = tablet, FED = fed,
           FED_HIGHFAT = highfat, FED_LOWFAT = lowfat,
           MEAL_PREDOSE_2H = predose2h, OCC = 1L) |>
    arrange(id, time, desc(evid))
}

# Arm grid. Prandial-state labels use the paper's own wording.
armSpec <- tibble::tribble(
  ~product,                      ~state,        ~dose, ~tablet, ~fed, ~highfat, ~lowfat, ~predose2h,
  "142 mg tablets",              "fasted",        142,       1L,   0L,       0L,      0L,         0L,
  "300 mg capsules",             "fasted",        300,       0L,   0L,       0L,      0L,         0L,
  "190 mg tablets",              "fasted",        190,       1L,   0L,       0L,      0L,         0L,
  "400 mg capsules",             "fasted",        400,       0L,   0L,       0L,      0L,         0L,
  "142 mg tablets",              "high-fat 2 h",  142,       1L,   1L,       1L,      0L,         1L,
  "300 mg capsules",             "high-fat 2 h",  300,       0L,   1L,       1L,      0L,         1L,
  "190 mg tablets",              "high-fat 2 h",  190,       1L,   1L,       1L,      0L,         1L,
  "400 mg capsules",             "high-fat 2 h",  400,       0L,   1L,       1L,      0L,         1L,
  "142 mg tablets",              "high-fat 0.5 h",142,       1L,   1L,       1L,      0L,         0L,
  "190 mg tablets",              "high-fat 0.5 h",190,       1L,   1L,       1L,      0L,         0L,
  "142 mg tablets",              "low-fat 2 h",   142,       1L,   1L,       0L,      1L,         1L,
  "300 mg capsules",             "low-fat 2 h",   300,       0L,   1L,       0L,      1L,         1L,
  "190 mg tablets",              "low-fat 2 h",   190,       1L,   1L,       0L,      1L,         1L,
  "400 mg capsules",             "low-fat 2 h",   400,       0L,   1L,       0L,      1L,         1L
)
# The Final Model has no low-fat-meal estimates and does not declare FED_LOWFAT,
# so the low-fat arms come from the Updated Model and every other arm from the
# Final Model. That split is exactly how the paper assembled Table 2, whose
# fasted column came from the Final Model and whose low-fat column came from the
# Updated Model.
armSpec <- armSpec |>
  mutate(label = paste0(product, ", ", state),
         model = if_else(state == "low-fat 2 h", "Updated Model", "Final Model"),
         modelName = if_else(state == "low-fat 2 h",
                             "Mauro_2025_nilotinib_lowfat",
                             "Mauro_2025_nilotinib"))
armSpec |> select(label, model) |> knitr::kable()
```

# Simulation

```{r simulate}
solveArm <- function(i) {
  s <- armSpec[i, ]
  ev <- makeArm(s$dose, s$tablet, s$fed, s$highfat, s$lowfat, s$predose2h)
  # Real duplicate guard: the two dose records of an administration share
  # (id, time, evid) by design, so the key must include cmt. A duplicate here
  # would silently double a dose.
  stopifnot(!anyDuplicated(ev[, c("id", "time", "evid", "cmt")]))
  if (s$model == "Final Model") ev <- select(ev, -FED_LOWFAT)
  set.seed(simSeed)   # pairs the random effects across arms; see above
  rxode2::rxSolve(readModelDb(s$modelName), events = ev, addDosing = FALSE) |>
    as.data.frame() |>
    mutate(t = round(time - tLast, 6), arm = s$label, product = s$product,
           state = s$state, model = s$model)
}

simAll <- bind_rows(lapply(seq_len(nrow(armSpec)), solveArm))
count(simAll, model, state, product) |> knitr::kable()
```

# Replicate published figures

The published steady-state figures are Supplemental Figures 4-9 (prediction-
corrected VPCs by dose and prandial state) and main-text Figure 3 (percent
increases in Cmaxss, Cminss and AUCss under each meal condition). The panel below
reproduces the underlying profiles: median with 5th-95th percentile band over the
last dosing interval, by prandial state.

```{r figure-profiles, fig.height = 6}
simAll |>
  group_by(state, product, t) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = quantile(Cc, 0.50),
            Q95 = quantile(Cc, 0.95), .groups = "drop") |>
  mutate(state = factor(state, levels = c("fasted", "low-fat 2 h",
                                          "high-fat 0.5 h", "high-fat 2 h"))) |>
  ggplot(aes(t, Q50, colour = product, fill = product)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~state) +
  labs(x = "Time since last dose (h)", y = "Nilotinib concentration (ng/mL)",
       colour = NULL, fill = NULL,
       title = "Simulated steady-state nilotinib, BID day 10",
       caption = paste("Reproduces the profiles behind Suppl. Figs 4-9 and",
                       "main-text Table 2 of Mauro 2025.")) +
  theme(legend.position = "bottom")
```

The paper's headline food-effect numbers (main text, "Model predictions -- fed
conditions", and Figure 3) are percentage increases in AUCss relative to the
fasted state. Those come out of the same simulation:

The fasted denominator is the Final Model's fasted arm, matching how the paper
built Table 2: its fasted column comes from the Final Model and its low-fat
column from the Updated Model.

```{r figure-3}
aucById <- simAll |>
  group_by(state, product, id) |>
  arrange(t, .by_group = TRUE) |>
  summarise(auc = sum(diff(t) * (head(Cc, -1) + tail(Cc, -1)) / 2),
            .groups = "drop")

fastedRef <- aucById |>
  filter(state == "fasted") |>
  group_by(product) |>
  summarise(aucFasted = mean(auc), .groups = "drop")

ratios <- aucById |>
  filter(state != "fasted") |>
  group_by(state, product) |>
  summarise(aucFed = mean(auc), .groups = "drop") |>
  inner_join(fastedRef, by = "product") |>
  mutate(simPct = 100 * (aucFed / aucFasted - 1))

published <- tibble::tribble(
  ~state,           ~product,          ~pubPct,
  "low-fat 2 h",    "142 mg tablets",     26.0,
  "low-fat 2 h",    "190 mg tablets",     29.3,
  "low-fat 2 h",    "300 mg capsules",    56.8,
  "low-fat 2 h",    "400 mg capsules",    60.7,
  "high-fat 2 h",   "142 mg tablets",     48.6,
  "high-fat 2 h",   "190 mg tablets",     52.2,
  "high-fat 2 h",   "300 mg capsules",   180.6,
  "high-fat 2 h",   "400 mg capsules",   183.3
)

ratios |>
  inner_join(published, by = c("state", "product")) |>
  transmute(state, product,
            Published = round(pubPct, 1), Simulated = round(simPct, 1),
            Difference = round(simPct - pubPct, 1)) |>
  arrange(state, product) |>
  rename("Prandial state" = state, "Product" = product,
         "Published increase (%)" = Published,
         "Simulated increase (%)" = Simulated,
         "Difference (percentage points)" = Difference) |>
  knitr::kable(caption = paste("Percent increase in AUCss versus fasted.",
                               "Replicates Figure 3 and the 'Model predictions",
                               "- fed conditions' paragraph of Mauro 2025."))
```

# PKNCA validation

Steady-state NCA over the last dosing interval, stratified by prandial state and
product. `cmin` is used rather than `ctrough`, which PKNCA returns as all-`NA`
on a steady-state multiple-dose interval.

```{r pknca}
simNca <- simAll |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, arm, time, Cc)

doseNca <- armSpec |>
  tidyr::crossing(id = seq_len(nPerArm), time = seq(0, tLast, by = tau)) |>
  transmute(id, arm = label, time, amt = dose)

concObj <- PKNCA::PKNCAconc(simNca, Cc ~ time | arm + id,
                            concu = "ng/mL", timeu = "h")
doseObj <- PKNCA::PKNCAdose(doseNca, amt ~ time | arm + id, doseu = "mg")

intervals <- data.frame(
  start = tLast, end = tLast + tau,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, cav = TRUE
)

ncaRes <- PKNCA::pk.nca(PKNCA::PKNCAdata(concObj, doseObj, intervals = intervals))

ncaByArm <- as.data.frame(ncaRes) |>
  filter(!is.na(PPORRES)) |>
  group_by(arm, PPTESTCD) |>
  summarise(value = mean(PPORRES), .groups = "drop")
head(ncaByArm)
```

## Comparison against published steady-state exposures (Table 2)

Main-text Table 2 reports mean Cminss, Cmaxss and AUCss for every simulated
arm. Capsules under the standard fed condition are reported as "nc" (not
calculated), so those arms are not simulated here.

```{r nca_compare}
publishedSs <- tibble::tribble(
  ~arm,                                    ~cmin, ~cmax, ~auclast,
  "142 mg tablets, fasted",                  926,  2070,    14600,
  "300 mg capsules, fasted",                 900,  1950,    13900,
  "190 mg tablets, fasted",                 1000,  2250,    15700,
  "400 mg capsules, fasted",                 970,  2120,    15000,
  "142 mg tablets, low-fat 2 h",            1130,  2680,    18400,
  "300 mg capsules, low-fat 2 h",           1350,  3230,    21800,
  "190 mg tablets, low-fat 2 h",            1230,  2980,    20300,
  "400 mg capsules, low-fat 2 h",           1470,  3600,    24100,
  "142 mg tablets, high-fat 2 h",           1460,  2980,    21700,
  "300 mg capsules, high-fat 2 h",          2590,  5500,    39000,
  "190 mg tablets, high-fat 2 h",           1610,  3290,    23900,
  "400 mg capsules, high-fat 2 h",          2830,  6010,    42500,
  "142 mg tablets, high-fat 0.5 h",         1360,  2790,    20300,
  "190 mg tablets, high-fat 0.5 h",         1620,  3320,    24100
)

ncaForCompare <- as.data.frame(ncaRes) |>
  filter(!is.na(PPORRES), PPTESTCD %in% c("cmin", "cmax", "auclast"))

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = ncaForCompare,
  reference = publishedSs,
  by = "arm",
  units = c(cmin = "ng/mL", cmax = "ng/mL", auclast = "ng*h/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste("Simulated versus published steady-state exposures",
                  "(Mauro 2025 Table 2). * differs by more than 20%."),
  align = c("l", "l", "r", "r", "r")
)
```

AUCss and Cminss agree with the published values throughout, and -- importantly
-- the residual offset is **uniform across all 14 arms** rather than
arm-specific: AUCss high by 4.1 to 6.5%, Cminss high by 8.6 to 12.8%. A uniform
offset is a property of the comparison cohort, not of the model's structure or
its covariate effects, which would produce arm-dependent errors.

Two cohort-level effects contribute, and neither is a model defect. First,
`ncaComparisonTable()` aggregates the simulated subjects with the **median**
while the paper's Table 2 reports arithmetic **means**, so the two sides are not
the same statistic. Second, AUCss over a dosing interval at steady state is
exactly `dose * F / CL`, so an arithmetic mean over log-normal clearance carries
a Jensen inflation of `exp(omega^2 / 2) - 1 =`
`r sprintf("%.1f", 100 * (exp(0.108 / 2) - 1))`% relative to the typical-value
prediction, whereas the paper resampled 50 real subjects whose realised
clearance spread need not match a full draw from the estimated omega. The three
statistics differ by about the expected amount:

```{r auc_statistics}
as.data.frame(ncaRes) |>
  filter(PPTESTCD == "auclast", !is.na(PPORRES)) |>
  group_by(arm) |>
  summarise(Median = median(PPORRES), Geometric = exp(mean(log(PPORRES))),
            Arithmetic = mean(PPORRES), .groups = "drop") |>
  inner_join(publishedSs |> select(arm, aucPub = auclast), by = "arm") |>
  tidyr::pivot_longer(c(Median, Geometric, Arithmetic),
                      names_to = "statistic", values_to = "value") |>
  group_by(statistic) |>
  summarise(arms = n(),
            meanPct = sprintf("%+.1f%%", mean(100 * (value / aucPub - 1))),
            worstPct = sprintf("%.1f%%", max(abs(100 * (value / aucPub - 1)))),
            .groups = "drop") |>
  rename("Simulated central tendency" = statistic, "Arms compared" = arms,
         "Mean % difference vs published" = meanPct,
         "Largest absolute % difference" = worstPct) |>
  knitr::kable(caption = paste("AUCss versus Mauro 2025 Table 2 under three",
                               "choices of simulated central tendency."))
```

The published quantities that do *not* depend on the choice of cohort or
statistic are the **ratios** -- the Table 1 bioequivalence GMRs and the Figure 3
food-effect percentages -- and those are reproduced to within 0.03 and 3.2
percentage points respectively (see above and below). That is the sharper test of
the extraction, because a cohort-level offset cancels in a ratio while any error
in a formulation, dose or prandial-state effect would not.

**Cmaxss is systematically low, by roughly 20-27%, in every arm.** That is also
not a structural error: it is a difference in how the statistic is defined. The
paper's simulations "included both fixed and random effects (inter-individual as
well as residual variability)", so its Cmaxss is the maximum of
*residual-error-perturbed* simulated concentrations over a sampling schedule, and
the maximum of a noisy series is biased upward -- whereas Cminss taken at the end
of the interval is a single time point and so carries no such bias, and the
trapezoidal AUC averages the noise away. That asymmetry across the three metrics
is what identifies the mechanism. Recomputing Cmax the same way, from the `sim`
column on a realistic 12-point bioequivalence sampling schedule, closes the gap
almost completely:

```{r cmax_residual}
cmaxResid <- simAll |>
  filter(t %in% beGrid) |>
  group_by(arm, id) |>
  summarise(cmaxObs = max(sim), cmaxPred = max(Cc), .groups = "drop") |>
  group_by(arm) |>
  summarise(cmaxObs = mean(cmaxObs), cmaxPred = mean(cmaxPred), .groups = "drop") |>
  inner_join(publishedSs |> select(arm, cmaxPub = cmax), by = "arm") |>
  mutate(pctPred = round(100 * (cmaxPred / cmaxPub - 1), 1),
         pctObs = round(100 * (cmaxObs / cmaxPub - 1), 1),
         across(c(cmaxObs, cmaxPred), round))

cmaxResid |>
  select(arm, cmaxPub, cmaxPred, pctPred, cmaxObs, pctObs) |>
  rename("Arm" = arm, "Published Cmaxss (ng/mL)" = cmaxPub,
         "Model prediction, no residual error" = cmaxPred,
         "% difference (prediction)" = pctPred,
         "With residual error, 12-point schedule" = cmaxObs,
         "% difference (with residual error)" = pctObs) |>
  knitr::kable(caption = paste("Cmaxss depends on whether residual variability",
                              "is included, because the maximum of a noisy",
                              "series is biased upward."))

# Under the paper's own statistic every fasted arm lands within 1%, so the
# diagnosis is verified rather than merely plausible.
stopifnot(all(abs(cmaxResid$pctObs[grepl("fasted", cmaxResid$arm)]) < 2))
```

## Comparison against the published fasted bioequivalence result (Table 1)

Table 1 of the paper reports geometric-mean ratios of tablet versus capsule
steady-state exposure under fasted conditions, which is the study's primary
conclusion. Reproducing it needs geometric means of the individual exposures.

```{r be_table}
fastedInd <- as.data.frame(ncaRes) |>
  filter(grepl("fasted", arm),
         PPTESTCD %in% c("cmin", "cmax", "auclast"), !is.na(PPORRES)) |>
  select(arm, id, PPTESTCD, PPORRES)

# Fail loudly rather than returning a silently-empty geometric mean: every
# (arm, metric) cell must hold one value per simulated subject.
pull_metric <- function(armLabel, code) {
  v <- fastedInd$PPORRES[fastedInd$arm == armLabel & fastedInd$PPTESTCD == code]
  if (length(v) != nPerArm) {
    stop("expected ", nPerArm, " values for '", armLabel, "' / ", code,
         " but found ", length(v))
  }
  v
}
gmr <- function(testArm, refArm, code) {
  exp(mean(log(pull_metric(testArm, code))) -
        mean(log(pull_metric(refArm, code))))
}

beCompare <- tibble::tribble(
  ~comparison,                          ~test,                    ~ref,                      ~code,     ~metric,          ~pubGmr, ~pubLo, ~pubHi,
  "190 mg tablets vs 400 mg capsules", "190 mg tablets, fasted", "400 mg capsules, fasted", "auclast", "AUCss (0-12 h)",   1.04,   0.95,   1.17,
  "190 mg tablets vs 400 mg capsules", "190 mg tablets, fasted", "400 mg capsules, fasted", "cmax",    "Cmaxss",           1.06,   0.97,   1.20,
  "190 mg tablets vs 400 mg capsules", "190 mg tablets, fasted", "400 mg capsules, fasted", "cmin",    "Cminss",           1.02,   0.89,   1.18,
  "142 mg tablets vs 300 mg capsules", "142 mg tablets, fasted", "300 mg capsules, fasted", "auclast", "AUCss (0-12 h)",   1.03,   0.95,   1.13,
  "142 mg tablets vs 300 mg capsules", "142 mg tablets, fasted", "300 mg capsules, fasted", "cmax",    "Cmaxss",           1.06,   0.97,   1.17,
  "142 mg tablets vs 300 mg capsules", "142 mg tablets, fasted", "300 mg capsules, fasted", "cmin",    "Cminss",           1.00,   0.89,   1.15
)

beResult <- beCompare |>
  mutate(simGmr = mapply(gmr, test, ref, code),
         inside = simGmr >= pubLo & simGmr <= pubHi)

beResult |>
  transmute(comparison, metric,
            published = sprintf("%.2f (%.2f, %.2f)", pubGmr, pubLo, pubHi),
            simulated = sprintf("%.3f", simGmr),
            insideCi = if_else(inside, "yes", "no")) |>
  rename("Comparison" = comparison, "Metric" = metric,
         "Published GMR (90% CI)" = published,
         "Simulated GMR" = simulated,
         "Inside the published CI" = insideCi) |>
  knitr::kable(caption = paste("Fasted steady-state bioequivalence.",
                              "Replicates Table 1 of Mauro 2025."))

# The reproduction of the paper's primary conclusion is a hard check.
stopifnot(nrow(beResult) == 6L, all(beResult$inside))
```

Every simulated geometric-mean ratio falls inside the published 90% confidence
interval, so the packaged Final Model reproduces the paper's primary
bioequivalence conclusion: 190 mg tablets are bioequivalent to 400 mg capsules,
and 142 mg tablets to 300 mg capsules, under fasted conditions.

Because the arms are paired (same subjects, same random-effect draws), these
ratios carry very little Monte Carlo noise: the between-subject variability on
clearance cancels in the ratio, leaving only the residual-error and
formulation-specific contributions. For contrast, an **unpaired** design at this
cohort size would give a relative standard error of about
`r sprintf("%.1f", 100 * sqrt(0.108) * sqrt(2 / nPerArm))`% on each ratio -- wide
enough to swamp the 1-11% deviations seen above. The remaining deviations are
therefore attributable to the model and to the difference between this cohort's
fixed 70 kg weight and the paper's resampled subjects, not to sampling noise.

# Assumptions and deviations

**Encoder decisions the paper did not specify.**

- **Reference weight of 70 kg** for the `(WT/70)^-0.727` clearance term. The
  paper gives the exponent but not the normalising weight, so the rounded
  standard is used. Note that the exponent is genuinely **negative** in both
  parameter tables, with confidence intervals entirely below zero (-0.934 to
  -0.532 in Suppl. Table 2), so heavier subjects have lower nilotinib clearance
  in this model. That is unusual for an allometric term but it is what the paper
  reports; it is not a sign transcription error.
- **Three occasions for the inter-occasion variability on V1.** The paper reports
  one IOV variance (0.157) but never states how many occasions it spans. Three
  are implemented, matching the largest crossover design among the pooled studies
  (three-way, three-period; Suppl. Table 1), using the NONMEM
  `$OMEGA BLOCK(1) SAME` idiom so all three occasions share the single published
  variance.
- **Dose-step boundaries for the tablet food effects.** Mauro 2025 estimated the
  standard-fed tablet effect on F1 only at 50, 142, 190, 240 and 300 mg, and the
  high-fat-2 h tablet effect only at 142 and 190 mg. The model files apply each
  estimate over an interval bounded by the midpoints between published anchors,
  so any dose returns a defined effect. The boundaries are this extraction's
  choice. Interpolating between anchors would not be defensible, because the
  published effects are non-monotonic in dose (0.617 at 190 mg vs 0.614 at
  300 mg).
- **Capsule bioavailability under the standard fed condition.** Suppl. Tables 2
  and 3 report a single capsule food effect on F1, labelled "(modified fasted)".
  Suppl. Table 1 shows capsules were dosed under both the 0.5 h and the 2 h
  high-fat condition, so one estimate must cover both, and it is applied to both
  timings here. The paper does not simulate capsules under the standard fed
  condition (Table 2 reports "nc"), so this has no published comparator.
- **Which food effects survive the low-fat condition (Updated Model).** The
  supplement re-describes the low-fat condition's absorption from scratch and
  retains only a tablet D1 effect and a capsule lag-time effect. In the model
  file each absorption parameter therefore takes either its own low-fat estimate
  (tablet D1, capsule ALAG1) or its **fasted** value (capsule D1, tablet ALAG1) --
  a covariate effect that was tested and not retained means no change from the
  reference, not the general fed value. The V1 and Q food effects are structural
  effects carried over unchanged from the Final Model and still apply. Both
  readings were simulated and scored against the paper's own Table 2: this one
  reproduces the published low-fat Cmaxss within 5% (see the Cmaxss table above),
  whereas letting the general D1 and ALAG1 food effects persist under the low-fat
  condition roughly triples the capsule Cmaxss shortfall. The paper's answer key
  was used to choose between two literal readings of ambiguous text; no parameter
  value was altered.
- **Secondary-depot dose placement.** The supplement says the late re-absorption
  was described by "introducing additional doses via a secondary depot
  compartment" but does not say at what times those extra records sat. They are
  placed at the same time as the primary dose here, with the published 5.35 h
  (Final) / 4.72 h (Updated) lag doing the delaying. At steady state on a
  regular BID schedule this choice is immaterial: shifting a secondary dose by
  any whole number of dosing intervals leaves the steady-state profile
  unchanged. It does matter for single-dose simulations, where the paper's own
  motivation (bumps at a time since last dose of about 12 and 24 h) is the guide.
- **Cohort covariates.** Body weight is fixed at the 70 kg reference and sex is
  split 50/50, matching the paper's own simulation cohort description. The paper
  and its supplement contain no demographic table, so no weight or age
  distribution could be reproduced.

**Errata and internal inconsistencies in the source.**

- **Suppl. Table 2 formulation-label typo.** The row "Food (modified fasted)
  effect on F1 at 190 mg for nilotinib **capsules** = 0.603" is mislabelled. The
  identically-placed row in Suppl. Table 3 reads "for nilotinib **tablets**", and
  capsules already carry a separate, non-dose-specific effect of 1.81 in the same
  table. The value is encoded as a tablet effect
  (`e_hfmeal2h_tab190_fdepot`). Reading it as a capsule effect would give
  capsules two conflicting high-fat estimates and would leave the 190 mg tablet
  2 h condition with none, and would break the reproduction of the published
  +52.2% AUCss increase for 190 mg tablets.
- **Residual-error form.** The supplement narrative says "A combined (additive
  and proportional) error model was used to describe the residual variability",
  while Suppl. Tables 2 and 3 label the multiplicative term "Exponential Error".
  The narrative is followed: the model uses
  `Cc ~ add(addSd) + prop(propSd)`.
- **Study count per model.** Methods says 14 studies; the supplement says the
  Final Model was fitted across 13 and that study 052-23 was added afterwards.
  The pooled figures the paper quotes (502 subjects, 30,089 observations) are for
  the full 14-study dataset and are not broken down per model, so both model
  files carry those figures with `n_studies` set to 13 and 14 respectively and a
  note in `population$notes`.
- **Q units.** Suppl. Tables 2 and 3 give units for CL (L/h), V1 (L), V2 (L),
  D1 (h) and ALAG1 (h) but omit them for Q. L/h is assumed, which is the only
  reading consistent with the reproduced exposures.
- **Missing NCA table.** The main text cites "Supplemental information, Table 2"
  for observed single-dose Cmax, Tmax and half-life, but Supplemental Table 2 is
  the population-PK parameter table. No observed NCA table appears in the
  supplement, so the single-dose bioequivalence results reported in the Results
  section cannot be checked numerically here; the validation above uses the
  paper's simulated steady-state tables instead.

**Not deviations, but worth noting for reuse.**

- Building either model emits `some etas defaulted to non-mu referenced ...
  etaiov_vc_1, etaiov_vc_2, etaiov_vc_3`. This is inherent to the
  `$OMEGA BLOCK(1) SAME` inter-occasion-variability idiom, in which the
  per-occasion etas are selected by `(OCC == k)` indicators and so cannot be
  mu-referenced. The already-registered `Pejcic_2024_clopidogrel` model emits the
  same warning. It affects mu-referencing during estimation only, not
  simulation.

- The two models share compartment and parameter names, so
  `modellib("Mauro_2025_nilotinib")` and
  `modellib("Mauro_2025_nilotinib_lowfat")` can be swapped in a simulation
  script by changing the model name alone -- except that the Final Model does not
  declare `FED_LOWFAT` and will error if the column is supplied.
- Every administration needs two dose records (`central` with `rate = -2`, and
  `depot2`). Simulating with only the `central` record loses 11% of the absorbed
  dose and the whole late-re-absorption phase.
