---
title: "Dostarlimab (Melhem 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dostarlimab (Melhem 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Melhem M, Hanze E, Lu Z, Venkatakrishnan K, Gupta N, Vugmeyster Y.
  Population pharmacokinetics and exposure-response of anti-programmed
  cell death protein-1 monoclonal antibody dostarlimab in advanced solid
  tumours. *Br J Clin Pharmacol.* 2022;88(9):4142-4154.
  doi:[10.1111/bcp.15339](https://doi.org/10.1111/bcp.15339)
* Description: Two-compartment population PK model for dostarlimab
  (anti-PD-1 IgG4) with time-dependent (sigmoid I_max) clearance in
  adults with advanced solid tumours.
* Modality: Therapeutic monoclonal antibody (IgG4 hinge-stabilized), IV
  infusion.

Dostarlimab is an anti-PD-1 monoclonal antibody. The Melhem 2022 paper
developed a population PK model from the Phase 1 GARNET trial
(NCT02715284) supporting the recommended therapeutic dose regimen of
**500 mg IV Q3W for four cycles followed by 1000 mg IV Q6W**.

Structure: linear two-compartment IV model with **time-dependent
clearance** via a sigmoid I_max function of time since first dose:

\[
  \mathrm{CL}(t) = \mathrm{CL}_{\text{base}} \cdot
  \exp\!\left(I_{\max}\cdot \dfrac{t^{\mathrm{Hill}}}
  {T_{50}^{\mathrm{Hill}} + t^{\mathrm{Hill}}}\right)
\]

with \(I_{\max} = -0.161\) (a fractional decrease in log-CL at
\(t \gg T_{50}\); \(1 - \exp(I_{\max}) = 0.149\), matching the paper's
"maximum reduction of 14.9%"), \(T_{50} = 108\) days, and
\(\mathrm{Hill} = 5.29\). Allometric weight scaling uses reference weight
70 kg (exponent 0.470 on CL, 0.419 on Vc and Vp).

# Population

The final-model population was 546 patients with **4783 PK observations**
(after exclusion of 21 records with absolute conditional weighted residual
> 5) drawn from a single multinational Phase 1 trial across multiple
expansion cohorts (Melhem 2022 Tables 1-2):

- 23.4% mismatch repair-deficient (dMMR/MSI-H) endometrial cancer (EC),
- 29.3% mismatch repair-proficient (MMRp/MSS) EC,
- 12.3% non-small-cell lung cancer (NSCLC),
- 28.8% non-EC dMMR/MSI-H or POLE-mutant tumours,
- and an ovarian-cancer cohort.

Demographics: median age 64 years (range 24-86), 77.3% female, median
weight 71.4 kg (range 34-182), median albumin 39 g/L, median ALT 17 U/L.
Hepatic and renal impairment were present at baseline (mild-to-moderate
in 11% and 61% of patients respectively) but did not affect dostarlimab
PK. ADAs ever-positive in 18.5% of patients but had no impact on
clearance.

The reference patient used in the published forest plot (Methods,
*PopPK model development*, last paragraph) is female, 70 kg, 64 years
old, with baseline albumin 39 g/L and ALT 18 U/L.

The same metadata is available programmatically via
`readModelDb("Melhem_2022_dostarlimab")$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Melhem_2022_dostarlimab.R`.
The table below collects them in one place for review.

| Parameter (model name)     | Value           | Source                                            |
|----------------------------|-----------------|---------------------------------------------------|
| `lcl` (CL_base, L/day)     | log(0.179)      | Melhem 2022 Table 3, CL                           |
| `lvc` (Vc_base, L)         | log(2.98)       | Melhem 2022 Table 3, Vc                           |
| `lq` (Q, L/day)            | log(0.547)      | Melhem 2022 Table 3, Q                            |
| `lvp` (Vp, L)              | log(2.10)       | Melhem 2022 Table 3, Vp                           |
| `lImax` (log\|I_max\|)     | log(0.161)      | Melhem 2022 Table 3, I_max = -0.161               |
| `lt50` (T50, days)         | log(108)        | Melhem 2022 Table 3, T50                          |
| `lhill` (Hill)             | log(5.29)       | Melhem 2022 Table 3, Hill                         |
| `e_wt_cl`                  | 0.470           | Melhem 2022 Table 3, Effect of WT on CL           |
| `e_wt_vc_vp`               | 0.419           | Melhem 2022 Table 3, Effect of WT on Vc and Vp    |
| `e_age_cl`                 | -0.227          | Melhem 2022 Table 3, Effect of age on CL          |
| `e_alb_cl`                 | -1.01           | Melhem 2022 Table 3, Effect of ALB on CL          |
| `e_alt_cl`                 | -0.0585         | Melhem 2022 Table 3, Effect of ALT on CL          |
| `e_alb_vc`                 | -0.153          | Melhem 2022 Table 3, Effect of ALB on Vc          |
| `e_sex_cl`                 | 0.165           | Melhem 2022 Table 3, Effect of male on CL         |
| `e_sex_vc`                 | 0.162           | Melhem 2022 Table 3, Effect of male on Vc         |
| IIV block `etalcl + etalvc`| c(0.0551, 0.0210, 0.0258) | Melhem 2022 Table 3 (omega^2 and covariance) |
| `etalImax`                 | 0.537           | Melhem 2022 Table 3, omega^2 I_max                |
| `propSd`                   | 0.133           | Melhem 2022 Table 3, proportional residual        |
| `addSd` (mg/L)             | 2.79            | Melhem 2022 Table 3, additive residual            |

Equations:

- `d/dt(central)` and `d/dt(peripheral1)`: standard two-compartment IV
  micro-constant form (kel = cl/vc, k12 = q/vc, k21 = q/vp). Source:
  Melhem 2022 Results, equation block ("The final PopPK model is
  mathematically described by the following equations").
- Time-dependent CL via exp-Hill: same equation block.
- Covariate effects on CL and Vc: power form (cov / ref)^theta for
  continuous covariates, multiplicative (1 + theta * SEXM) for sex.

# Virtual cohort

Original observed data are not publicly available. The simulations below
use a virtual cohort whose covariate distributions approximate the
published trial demographics (Melhem 2022 Table 2).

```{r cohort}
set.seed(2022)
n_subj <- 200

cohort <- tibble(
  ID   = seq_len(n_subj),
  WT   = pmin(pmax(rlnorm(n_subj, log(71.4), 0.27), 34),  182),  # Table 2: median 71.4 kg, range 34-182
  AGE  = pmin(pmax(rnorm(n_subj,  62.5, 11.0),       24),  86),  # Table 2: mean 62.5, range 24-86
  ALB  = pmin(pmax(rnorm(n_subj,  38.2, 5.1),        19,        51)),  # Table 2: mean 38.2 g/L, range 19-51
  ALT  = pmin(pmax(rlnorm(n_subj, log(17), 0.5),     2.9),      120),  # Table 2: median 17, range 2.9-120
  SEXF = rbinom(n_subj, 1, 0.773)                                       # Table 2: 77.3% female
)
```

The recommended therapeutic dose (RTD) is **500 mg IV Q3W for four
cycles followed by 1000 mg IV Q6W**. The simulation runs for one year
of dosing.

```{r events}
# RTD schedule (days from first dose).
loading_doses_d <- seq(0, by = 21, length.out = 4)              # 500 mg Q3W x 4
maint_start_d   <- max(loading_doses_d) + 21                    # first 1000 mg dose 21 d after the 4th 500 mg
maint_doses_d   <- seq(maint_start_d, by = 42, length.out = 12) # 1000 mg Q6W

obs_times_d <- sort(unique(c(seq(0, 365, by = 1), loading_doses_d, maint_doses_d)))

build_events <- function(pop) {
  d_load <- pop |>
    tidyr::crossing(TIME = loading_doses_d) |>
    mutate(AMT = 500, EVID = 1, CMT = "central", DUR = 0.5 / 24, DV = NA_real_,
           treatment = "RTD 500 mg Q3W x4 -> 1000 mg Q6W")
  d_maint <- pop |>
    tidyr::crossing(TIME = maint_doses_d) |>
    mutate(AMT = 1000, EVID = 1, CMT = "central", DUR = 0.5 / 24, DV = NA_real_,
           treatment = "RTD 500 mg Q3W x4 -> 1000 mg Q6W")
  d_obs <- pop |>
    tidyr::crossing(TIME = obs_times_d) |>
    mutate(AMT = NA_real_, EVID = 0, CMT = "central", DUR = NA_real_,
           DV = NA_real_, treatment = "RTD 500 mg Q3W x4 -> 1000 mg Q6W")
  dplyr::bind_rows(d_load, d_maint, d_obs) |>
    dplyr::arrange(ID, TIME, dplyr::desc(EVID)) |>
    as.data.frame()
}

events <- build_events(cohort)
```

# Simulation

```{r simulate}
mod <- readModelDb("Melhem_2022_dostarlimab")
sim <- rxSolve(mod, events = events, returnType = "data.frame")
```

# Concentration-time profiles

Median and 5-95th percentile envelope across the virtual cohort under
the recommended therapeutic dose:

```{r pk-plot, fig.width = 7, fig.height = 5}
sim_summary <- sim |>
  dplyr::filter(time > 0) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    median = stats::median(Cc, na.rm = TRUE),
    lo     = stats::quantile(Cc, 0.05, na.rm = TRUE),
    hi     = stats::quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(sim_summary, aes(time, median)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, fill = "steelblue") +
  geom_line(linewidth = 1, colour = "steelblue") +
  labs(
    x = "Time (days)",
    y = "Dostarlimab serum concentration (mg/L)",
    title = "Simulated dostarlimab PK at the recommended therapeutic dose",
    subtitle = paste0("Median and 90% prediction interval (N = ",
                      n_subj, " virtual patients)"),
    caption = "Model: Melhem 2022 Br J Clin Pharmacol 88(9):4142-4154"
  ) +
  theme_bw()
```

# Time-dependent clearance

Melhem 2022 reports time-dependent clearance with a sigmoid I_max function
of time since first dose. The typical-value CL profile below reproduces
the time course at the reference patient (deterministic, etas = 0):

```{r time-dep-cl, fig.width = 6, fig.height = 4}
t_grid <- seq(0, 365, by = 5)
events_cl <- data.frame(
  ID   = 1, WT = 70, AGE = 64, ALB = 39, ALT = 18, SEXF = 1,
  TIME = c(0, t_grid),
  AMT  = c(500, rep(NA_real_, length(t_grid))),
  EVID = c(1,  rep(0, length(t_grid))),
  CMT  = "central",
  DUR  = c(0.5 / 24, rep(NA_real_, length(t_grid))),
  DV   = NA_real_
)
sim_cl <- rxSolve(
  mod, events = events_cl,
  params = c(WT = 70, AGE = 64, ALB = 39, ALT = 18, SEXF = 1,
             etalcl = 0, etalvc = 0, etalImax = 0),
  omega  = NA,
  returnType = "data.frame"
)
sim_cl <- sim_cl[sim_cl$time > 0, ]

ggplot(sim_cl, aes(time, cl / cl_base)) +
  geom_line(linewidth = 1, colour = "steelblue") +
  geom_hline(yintercept = exp(-0.161), linetype = "dashed", colour = "grey40") +
  labs(
    x = "Time since first dose (days)",
    y = "CL(t) / CL(0)",
    title = "Time-dependent clearance in Melhem 2022",
    subtitle = "Reference patient; asymptotes at exp(I_max) = 0.851 (14.9% reduction)"
  ) +
  theme_bw()
```

# PKNCA validation

NCA computed over the third 21-day dosing interval (2nd-3rd 500 mg dose
window, days 21-42 after first dose) for the simulated reference cohort:

```{r pknca}
interval_start <- 21
interval_end   <- 42

sim_nca <- sim |>
  dplyr::filter(!is.na(Cc),
                time >= interval_start,
                time <= interval_end) |>
  dplyr::mutate(time_rel  = time - interval_start,
                treatment = "RTD 500 mg Q3W x4 -> 1000 mg Q6W") |>
  dplyr::select(id, treatment, time_rel, Cc)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time_rel | treatment + id)

dose_df <- data.frame(
  id        = cohort$ID,
  treatment = "RTD 500 mg Q3W x4 -> 1000 mg Q6W",
  time_rel  = 0,
  amt       = 500
)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time_rel | treatment + id)

intervals <- data.frame(
  start     = 0,
  end       = 21,
  cmax      = TRUE,
  tmax      = TRUE,
  auclast   = TRUE,
  half.life = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- PKNCA::pk.nca(nca_data)
knitr::kable(summary(nca_res),
             caption = "Simulated NCA parameters (3rd dosing interval, days 21-42 after first dose)")
```

## Comparison against published NCA

Melhem 2022 Table 4 reports per-patient cycle-1 (Day 21) exposure summary
statistics for the efficacy ER subset (N = 362, all from the RTD Part 2B
arm):

| Metric        | Published (Table 4) | Notes |
|---------------|---------------------|-------|
| Cmin (mg/L)   | mean 41.3 (SD 10.5), range 13.7-70.2  | Cycle 1 trough, RTD |
| Cmax (mg/L)   | mean 177 (SD 32.8), range 103-284     | Cycle 1 peak, RTD   |
| AUC (mg*h/L)  | mean 34 400 (SD 6090), range 19 100-51 000 | Cycle 1 over 21 d, RTD |

Note the published AUC is reported in **mg*h/L** but is computed over the
21-day cycle 1 interval; the simulated `auclast` from PKNCA in the table
above is in **mg*day/L** because the simulation time variable is in days.
Multiplying simulated `auclast` by 24 converts to mg*h/L for comparison.
Simulated cycle-1 Cmax in the 150-200 mg/L range (single 500 mg dose into
~3 L Vc, with a small fraction redistributed peripherally by day 1) is
consistent with the published Table 4 mean of 177 mg/L.

# Assumptions and deviations

- **Residual error convention.** Melhem 2022 Table 3 reports the
  proportional residual variability as 0.133 (unitless) and the additive
  residual variability as 2.79 mg/L. The additive entry carries the
  units of the observation (mg/L, not mg^2/L^2), indicating that the
  table column is the *standard deviation*, not the variance. Both
  entries are therefore used directly as `propSd` and `addSd` in
  nlmixr2's combined `add(addSd) + prop(propSd)` form. A 13.3% CV is
  consistent with mAb assay residuals reported in comparable anti-PD-1
  papers.
- **Imax parameterization.** I_max is always negative in the source
  (-0.161). To keep every individual I_max strictly negative under
  log-normal IIV, the model stores log|I_max| and applies the negative
  sign in the `model()` block: `Imax_i <- -exp(lImax + etalImax)`. The
  table's omega^2_Imax = 0.537 maps to omega = sqrt(0.537) = 0.733,
  which the paper labels "73.3% CV" — i.e., omega is reported as a CV
  without the log-normal CV = sqrt(exp(omega^2) - 1) correction. The
  packaged model uses 0.537 directly as the variance of `etalImax`,
  preserving the paper's table value at the cost of a ~10 percentage-
  point divergence between "labelled CV" and "true log-normal CV"; the
  same parameterization is used by `Masters_2022_avelumab.R`.
- **Albumin units typo in Methods.** The Methods text describes the
  reference patient with "ALB 39 g dL-1", but Table 2 lists albumin in
  g/L (median 39, range 19-51). 39 g/dL is physiologically impossible
  (normal serum albumin is 35-50 g/L = 3.5-5 g/dL). The model uses g/L
  as the canonical unit and 39 g/L as the reference, consistent with
  Table 2 and with the same range reported elsewhere in the paper.
- **Time variable.** The packaged model uses rxode2's `t` (simulation
  time in days) for the time-dependent CL term. Simulations must
  therefore start at t = 0 with the first dose so that the time-on-CL
  profile aligns with the source's "time since first dose" convention.
- **Time-varying covariates.** ALB and ALT enter the published model as
  time-varying covariates. The simulations in this vignette hold each
  subject's ALB and ALT fixed at their baseline value over the dosing
  window. Modelling on-treatment ALB/ALT trajectories is out of scope
  for the packaged structural model; users with longitudinal lab data
  can pass time-varying ALB / ALT columns to `rxSolve` directly via the
  event table.
- **Race / ethnicity in the cohort.** The simulated cohort does not
  include explicit race / ethnicity columns because race did not enter
  the final population PK model in Melhem 2022. The race distribution
  in the published analysis set (Table 2: 75.1% White, 3.5% Black,
  2.4% Asian, 16.3% Not reported) is recorded in the model's
  `population` metadata for users who need to stratify simulations.
- **Race covariates not declared.** No `RACE_*` covariate is registered
  in `covariateData` because the structural model has no race effect.
- **Erratum.** A 15 June 2022 publisher correction was issued for the
  copyright line only; no parameter, equation, or unit was changed.
