---
title: "Acetaminophen placental cotyledon perfusion (Mian 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Acetaminophen placental cotyledon perfusion (Mian 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
mod <- readModelDb("Mian_2020_acetaminophen_placentaperfusion")
ui <- rxode2::rxode(mod)
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1007/s40262-020-00861-7>
* Open-access full text: <https://europepmc.org/article/MED/PMC7329787>

# What this file is, and what it is not

Mian and colleagues built two kinds of model:

| # | Model | Implementation | In this file? |
|---|-------|----------------|---------------|
| 1 | **Ex vivo cotyledon perfusion model** (Section 2.5.1.2, Figure 3, Equations 5-8, Table 1) | MONOLIX | **yes** |
| 2 | Whole-body fetal-maternal PBPK for acetaminophen and its metabolites, with a fetal-liver compartment and three placental-transfer parameterisations (Table 2) | MoBi / PK-Sim | no |

Model 2 extends the authors' earlier pregnancy PBPK model (Mian et al.,
*Clin Pharmacokinet* 2020;59:97-110). Its tissue partition coefficients,
organ volumes and blood flows, enzyme reference concentrations and the
fetal-liver permeabilities and surface areas are all computed inside the
Open Systems Pharmacology software ("automatically estimated using scaling
approaches implemented in the software", Section 2.2) and are not printed in
either paper, so it cannot be reconstructed from the publications.

Model 1 is self-contained. It is four ODEs whose rate laws are printed as
Equations 5-8, whose volumes and flows are the stated experimental settings,
and whose three fitted constants are in Table 1. It is also the part of the
paper that transfers: `D_cot` and `K_f,m` are the quantities the authors
up-scaled into the whole-body model (Equation 9), and the ex vivo route was
the most accurate of the three placental-transfer approaches they compared
(Tables 3 and 4).

# Population

The model describes an isolated human term placenta, not a patient
population. One intact cotyledon per placenta was perfused in a recirculating
(closed-closed) dual circuit within 30 min of delivery, with the chorionic
artery and vein cannulated (Section 2.5.1.1). Acetaminophen 10 ug/mL -- the
steady-state concentration at clinical use -- was added to the maternal
reservoir. Maternal and fetal flows were 14 and 6 mL/min and the mean
reservoir volumes 280 and 284 mL. Samples were drawn from both reservoirs at
0, 3, 6, 10, 15, 20 and 30 min, every 15 min to 150 min, then every 30 min to
210 min. The number of placentas perfused is given in the earlier perfusion
report (reference 18 of the paper), not in this one.

# Source trace

| Quantity | Value | Source |
|----------|-------|--------|
| Maternal flow `Q_m` (`q_mat`) | 14 mL/min | Section 2.5.1.1 |
| Fetal flow `Q_f` (`q_fet`) | 6 mL/min | Section 2.5.1.1 |
| Maternal reservoir `V_m` (`v_mres`) | 280 mL | Section 2.5.1.1 |
| Fetal reservoir `V_f` (`v_fres`) | 284 mL | Section 2.5.1.1 |
| Maternal part of cotyledon `V_mp` (`v_mcot`) | 23 mL, fixed | Section 2.5.1.2 |
| Fetal part of cotyledon `V_fp` (`v_fcot`) | 35 mL, fixed | Section 2.5.1.2 |
| `D_cot` (`ldcot`) | 36 mL/min (81%) | Table 1 |
| `K_pe` (`lkpe`) | 0.0126 1/min (229%) | Table 1 |
| `K_f,m` (`lkfm`) | 0.737 (35%) | Table 1 |
| Maternal reservoir balance | Equation 5 | Section 2.5.1.2 |
| Maternal cotyledon balance | Equation 6 | Section 2.5.1.2 |
| Fetal cotyledon balance | Equation 7 | Section 2.5.1.2 |
| Fetal reservoir balance | Equation 8 | Section 2.5.1.2 |
| Initial maternal concentration | 10 ug/mL | Section 2.5.1.1 |
| Fitted maternal and fetal curves | read off the figure | Figure 4 |

# Simulation

The experiment is one dose of 10 ug/mL x 280 mL = 2800 ug into the maternal
reservoir, followed for 210 min.

```{r simulate}
dose_ug <- 10 * 280
tgrid <- sort(unique(c(seq(0, 210, by = 1), c(3, 6, 10, 15, 20, 30))))
ev <- et(amt = dose_ug, cmt = "maternal_reservoir") |>
  et(tgrid, cmt = "maternal_reservoir")
sim <- as.data.frame(rxSolve(mod, ev))
head(sim[, c("time", "Cmaternal", "Cfetal")])
```

# Replicate published figures

Figure 4 of the paper shows the model-predicted maternal and fetal reservoir
curves against the ex vivo observations. The two predicted curves were read
off the figure by the maintainers at the plotted time points (reading
precision about 0.1 ug/mL).

```{r fig4-digitised}
fig4 <- tibble::tribble(
  ~time_h, ~maternal, ~fetal,
  0.05,  8.75, 0.10,
  0.10,  8.40, 0.30,
  0.167, 8.00, 0.62,
  0.25,  7.55, 1.00,
  0.333, 7.20, 1.35,
  0.50,  6.60, 1.90,
  0.75,  5.90, 2.55,
  1.00,  5.40, 3.00,
  1.25,  5.05, 3.30,
  1.50,  4.75, 3.55,
  1.75,  4.58, 3.72,
  2.00,  4.45, 3.83,
  2.25,  4.35, 3.92,
  2.50,  4.28, 3.97,
  3.00,  4.20, 4.04,
  3.50,  4.15, 4.08
) |>
  mutate(time = round(time_h * 60))

ev_fig <- et(amt = dose_ug, cmt = "maternal_reservoir") |>
  et(sort(unique(fig4$time)), cmt = "maternal_reservoir")
sim_fig <- as.data.frame(rxSolve(mod, ev_fig)) |>
  select(time, Cmaternal, Cfetal)

cmp <- fig4 |>
  left_join(sim_fig, by = "time") |>
  mutate(
    pct_maternal = 100 * (Cmaternal - maternal) / maternal,
    pct_fetal = 100 * (Cfetal - fetal) / fetal
  )
cmp |>
  select(time, maternal, Cmaternal, pct_maternal, fetal, Cfetal, pct_fetal) |>
  dplyr::rename(
    "Time (min)" = time,
    "Maternal, Figure 4" = maternal,
    "Maternal, model" = Cmaternal,
    "Maternal diff (%)" = pct_maternal,
    "Fetal, Figure 4" = fetal,
    "Fetal, model" = Cfetal,
    "Fetal diff (%)" = pct_fetal
  ) |>
  knitr::kable(digits = 2)
```

```{r fig4-plot, fig.width = 7, fig.height = 4.5}
fig4_long <- fig4 |>
  select(time, maternal, fetal) |>
  pivot_longer(-time, names_to = "reservoir", values_to = "conc")
sim_long <- sim |>
  select(time, maternal = Cmaternal, fetal = Cfetal) |>
  pivot_longer(-time, names_to = "reservoir", values_to = "conc")
ggplot() +
  geom_line(data = sim_long, aes(time / 60, conc, colour = reservoir)) +
  geom_point(data = fig4_long, aes(time / 60, conc, colour = reservoir), shape = 1) +
  labs(
    x = "Time (h)", y = "Acetaminophen (ug/mL)", colour = "Reservoir",
    caption = "Replicates Figure 4 of Mian 2020. Lines: this model; circles: the paper's fitted curves read off the figure."
  ) +
  theme_bw()
```

The maternal curve is reproduced within 5% throughout. The fetal curve rises
faster than the published one early on -- 20-35% above it in the first
20 min, while the fetal concentration is still below 1.5 ug/mL, about 20% at
30 min and 14% at 1 h -- and converges on it from 2 h onward (within 1% at
210 min). The published curves are the
fit to pooled perfusions -- the figure's own maternal curve starts at
9.5 ug/mL rather than the nominal 10 -- whereas this file uses the stated
nominal settings, so an early-phase gap of this size is expected; it is not
tuned away.

```{r fig4-gates}
late <- cmp$time >= 60
stopifnot(
  # Maternal decline: a wrong flow, volume or D_cot moves this by tens of %.
  median(abs(cmp$pct_maternal)) < 5,
  max(abs(cmp$pct_maternal)) < 10,
  # Fetal rise: centre within 15%, and within 15% everywhere after 1 h.
  median(abs(cmp$pct_fetal[cmp$time >= 30])) < 15,
  max(abs(cmp$pct_fetal[late])) < 15
)
```

# Structural checks

## K_f,m acts on the cotyledon outflows, not on the diffusion step

Equations 5-8 place `K_f,m` on the flow terms leaving *both* halves of the
cotyledon (`Q_m x C_mp / K_f,m` and `Q_f x C_fp / K_f,m`), not on the
diffusion term between them. At equilibrium (and ignoring the small `K_pe`
loss) that makes the two reservoirs equal: Equation 5 gives
`C_mp = K_f,m x C_m`, Equation 8 gives `C_fp = K_f,m x C_f`, and the
diffusion term in Equations 6-7 vanishes only when `C_mp = C_fp`. The
Discussion instead describes `K_f,m` as "the maternal-to-fetal concentration
ratio at equilibrium", which would need the partition on the diffusion step,
`D_cot x (C_mp - C_fp / K_f,m)`, and would drive the fetal:maternal reservoir
ratio towards 0.737.

Figure 4 decides between the two. Its fitted curves end at 4.15 (maternal)
and 4.08 ug/mL (fetal), a ratio of 0.98, which the printed equations
reproduce and the Discussion's reading does not.

```{r kfm-reading}
alt <- rxode2({
  Cm <- mres / 280; Cmp <- mcot / 23; Cfp <- fcot / 35; Cf <- fres / 284
  d/dt(mres) <- 14 * Cmp - 14 * Cm
  d/dt(mcot) <- 14 * Cm - 14 * Cmp - 36 * (Cmp - Cfp / 0.737)
  d/dt(fcot) <- 6 * Cf - 6 * Cfp + 36 * (Cmp - Cfp / 0.737) - 0.0126 * 35 * Cfp
  d/dt(fres) <- 6 * Cfp - 6 * Cf
})
ev_end <- et(amt = dose_ug, cmt = "mres") |> et(c(210, 2000), cmt = "mres")
alt_end <- as.data.frame(rxSolve(alt, ev_end))
ev_end_mod <- et(amt = dose_ug, cmt = "maternal_reservoir") |>
  et(c(210, 2000), cmt = "maternal_reservoir")
mod_end <- as.data.frame(rxSolve(mod, ev_end_mod))

ratios <- tibble::tibble(
  Reading = c("Figure 4 (published fit)", "Equations 5-8 as printed", "Partition on the diffusion step"),
  `Fetal:maternal at 210 min` = c(
    4.08 / 4.15,
    mod_end$Cfetal[mod_end$time == 210] / mod_end$Cmaternal[mod_end$time == 210],
    alt_end$Cf[alt_end$time == 210] / alt_end$Cm[alt_end$time == 210]
  )
)
knitr::kable(ratios, digits = 3)

r_fig <- ratios[[2]][1]
stopifnot(
  abs(ratios[[2]][2] - r_fig) < 0.03,
  abs(ratios[[2]][3] - r_fig) > 0.15
)
```

## Mass balance and the closed-form equilibrium

With `K_pe = 0` the system is closed, so the total amount must stay at the
dose and every compartment must settle where the partitioning puts it:
reservoirs at `D / (V_m + V_f + K_f,m (V_mp + V_fp))` and both cotyledon
halves at `K_f,m` times that. Both sides of this check use the same
parameters, so a tight bound is correct.

```{r mass-balance}
ev_mb <- et(amt = dose_ug, cmt = "maternal_reservoir") |>
  et(c(10, 60, 210, 5000), cmt = "maternal_reservoir")
mb <- as.data.frame(rxSolve(mod, ev_mb, params = c(lkpe = log(1e-12)))) |>
  mutate(total = maternal_reservoir + maternal_cotyledon + fetal_cotyledon + fetal_reservoir)
c_eq <- dose_ug / (280 + 284 + 0.737 * (23 + 35))
mb_end <- mb[mb$time == 5000, ]
stopifnot(
  all(abs(mb$total - dose_ug) / dose_ug < 1e-6),
  abs(mb_end$Cmaternal / c_eq - 1) < 1e-4,
  abs(mb_end$Cfetal / c_eq - 1) < 1e-4,
  abs(mb_end$maternal_cotyledon / 23 / (0.737 * c_eq) - 1) < 1e-4
)
c(closed_form_ug_per_mL = c_eq, simulated_maternal = mb_end$Cmaternal)
```

## Placental elimination over the experiment

`K_pe` was estimated with a standard error of 229% (Table 1 labels the
column "residual standard error") and the authors called it "negligibly
small" (Section 3.2). Over the 210-min experiment it removes the following
share of the dose:

```{r kpe-loss}
end210 <- sim[sim$time == 210, ]
lost <- dose_ug - with(end210, maternal_reservoir + maternal_cotyledon + fetal_cotyledon + fetal_reservoir)
pct_lost <- 100 * lost / dose_ug
pct_lost
stopifnot(pct_lost > 5, pct_lost < 20)
```

About a tenth of the dose, so within the experiment `K_pe` is small but not
negligible; it accounts for the reservoirs ending near 4.1 ug/mL rather than
at the closed-system equilibrium of 4.6 ug/mL computed above. The authors'
judgement concerned the whole-body model, where `K_pe` was dropped.

## Up-scaling to the whole placenta (Equation 9)

Equation 9 scales the cotyledon clearance to the whole placenta by volume,
`D_pl = D_cot x V_pl / V_cot`. Table 2 gives `D_pl` = 403 mL/min from
`D_cot` = 36 mL/min and `V_cot` = 58 mL; the placental volume used is not
printed but follows as

```{r eq9}
v_pl <- 403 * 58 / 36
v_pl
stopifnot(v_pl > 400, v_pl < 900)
```

about 650 mL, a plausible term placental volume.

# NCA

The paper reports no non-compartmental parameters for the ex vivo experiment
(the Table 3 NCA values belong to the whole-body PBPK model against in vivo
cord-blood data), so there is no published NCA to compare with. For reference,
PKNCA summarises the simulated maternal-reservoir decline and the fetal
reservoir exposure over the 210-min experiment.

```{r pknca}
conc <- sim |>
  select(time, Cmaternal, Cfetal) |>
  pivot_longer(-time, names_to = "reservoir", values_to = "conc") |>
  mutate(id = 1L) |>
  filter(!is.na(conc))
dose_df <- tibble::tibble(
  id = 1L, time = 0, amt = dose_ug,
  reservoir = c("Cmaternal", "Cfetal")
)
conc_obj <- PKNCAconc(conc, conc ~ time | reservoir + id)
dose_obj <- PKNCAdose(dose_df, amt ~ time | reservoir + id)
intervals <- data.frame(start = 0, end = 210, cmax = TRUE, tmax = TRUE, auclast = TRUE)
nca <- pk.nca(PKNCAdata(conc_obj, dose_obj, intervals = intervals))
as.data.frame(nca$result) |>
  select(reservoir, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::rename(
    "Reservoir" = reservoir,
    "Cmax (ug/mL)" = cmax,
    "Tmax (min)" = tmax,
    "AUC0-210 (min*ug/mL)" = auclast
  ) |>
  knitr::kable(digits = 2)
```

# Assumptions and deviations

* **Only the ex vivo model is packaged.** The whole-body fetal-maternal PBPK
  model depends on partition coefficients, organ volumes and flows, enzyme
  reference concentrations and fetal-liver permeabilities that the Open
  Systems Pharmacology software computes and neither this paper nor its
  predecessor prints.
* **Equations 5-8 are used as printed.** The paper writes each balance as
  `dN/dt = (flux terms in N) / V`; the right-hand-side `N` are read as
  concentrations and the balances multiplied through by `V` to give amounts.
  `K_f,m` therefore divides the cotyledon concentration on both the maternal
  and the fetal outflow. That form makes the reservoirs equilibrate at a
  fetal:maternal ratio near 1, which disagrees with the Discussion's wording
  but matches the paper's own fitted curves in Figure 4 (see Structural
  checks).
* **Nominal starting concentration.** The simulation starts the maternal
  reservoir at the stated 10 ug/mL. Figure 4's fitted maternal curve starts
  near 9.5 ug/mL, the likely source of the early fetal difference.
* **K_pe is retained.** The authors left `K_pe` out of the whole-body PBPK
  model because it was small and imprecise, but it is part of the fitted ex
  vivo model and is kept here; set `lkpe` to a very small value to remove it.
* **No variability.** Between-placenta variability and the residual-error
  model were estimated (Section 2.5.1.2) but their final values are not
  reported, so the file carries typical values only and has no residual-error
  lines.
* **Metabolites.** Acetaminophen-glucuronide and -sulphate transfer were
  measured in the same perfusions (reference 18) but not modelled in this
  paper, and are not included.
* **Figure 4 values were read off the figure** and carry about 0.1 ug/mL of
  reading error; the observed perfusion concentrations in that figure are not
  used.
