---
title: "Teclistamab (Miao 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Teclistamab (Miao 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Miao_2023_teclistamab"))$reference`
* Description: `r rxode2::rxode(readModelDb("Miao_2023_teclistamab"))$description`
* Article: <https://doi.org/10.1007/s11523-023-00989-z>
* Supplement (Springer ESM, PDF): <https://static-content.springer.com/esm/art%3A10.1007%2Fs11523-023-00989-z/MediaObjects/11523_2023_989_MOESM1_ESM.pdf>

Teclistamab is a B-cell maturation antigen (BCMA) x CD3 bispecific antibody
approved for relapsed/refractory multiple myeloma (RRMM). Miao 2023 reports the
population pharmacokinetics from MajesTEC-1, the phase I/II first-in-human trial,
together with exposure-response analyses for efficacy and safety.

The PK model is a two-compartment model with first-order subcutaneous absorption
and **two parallel elimination pathways**: a time-independent clearance `CL1` and
a time-dependent clearance `CL2` that decays exponentially from the start of
treatment,

$$ CL(t) = CL_1 + CL_2 \cdot e^{-K_{DES}\,t}, \qquad t \text{ in days.} $$

Miao 2023 presents this as an empirical approximation of target-mediated drug
disposition coupled with a feedback loop from improving disease status onto
clearance. Quasi-steady-state TMDD models and models linking the time-dependent
clearance to time-varying soluble BCMA were both fitted and rejected (Discussion).

# Population

The analysis dataset held 4840 measurable serum teclistamab concentrations from
338 PK-evaluable patients with RRMM: 83 patients (1976 observations) dosed
intravenously and 255 patients (2864 observations) subcutaneously. Patients had a
median age of 64 years (range 24-84) and a median body weight of 74.3 kg (range
41.0-139); 44.4% were female and 83.4% were White. Baseline International Staging
System (ISS) stage was I in 48.8%, II in 32.2% and III in 17.5%; 51.2% had
IgG-secreting myeloma. Renal function was normal or mildly impaired in 75.2% and
hepatic function was normal in 88.2% (Miao 2023 Table 1, "Combined IV and SC
dosing" column). Only two of 321 ADA-evaluable patients developed anti-drug
antibodies.

The same information is available programmatically:

```{r population}
pop <- rxode2::rxode(readModelDb("Miao_2023_teclistamab"))$population
str(pop, max.level = 1)
```

# Source trace

Every `ini()` entry in `inst/modeldb/specificDrugs/Miao_2023_teclistamab.R`
carries an in-file comment naming its source location. They are collected here.

All values come from **Miao 2023 Table 2**, "Parameter estimates of teclistamab
for the final population PK model". The four covariate equations are the
footnotes a-d of that table; they were read from the MathML in the EuropePMC
JATS full text of PMC10518021, because the `pdftotext` and docling renderings of
footnote c lose the division bar and print `V1 = 4.13 x (BWT)^0.358` with a
stranded `74` on the following line.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl_exp_inf` (CL1) | 0.449 L/day | Table 2, RSE 8.87% |
| `lcl_exp_component` (CL2 at t = 0) | 0.547 L/day | Table 2, RSE 15.6% |
| `lcl_exp_kdes` (KDES) | 0.0292 1/day | Table 2, RSE 13.0% |
| `lvc` (V1) | 4.13 L | Table 2, RSE 4.40% |
| `lvp` (V2) | 1.34 L | Table 2, RSE 26.1% |
| `lq` (Q) | 0.0390 L/day | Table 2, RSE 55.5% |
| `lka` (Ka) | 0.133 1/day | Table 2, RSE 7.73% |
| `lfdepot` (F) | 0.718 | Table 2, RSE 7.38% |
| `e_wt_cl_exp_inf` | 0.704 | Table 2 footnote a, RSE 21.8% |
| `e_iss_ii_cl_exp_inf` | 1.31 | Table 2 footnote a, RSE 7.83% |
| `e_iss_iii_cl_exp_inf` | 1.67 | Table 2 footnote a, RSE 11.1% |
| `e_nigg_cl_exp_inf` | 0.689 | Table 2 footnote a, RSE 7.76% |
| `e_nigg_cl_exp_component` | 0.295 | Table 2 footnote b, RSE 21.6% |
| `e_wt_vc` | 0.358 | Table 2 footnote c, RSE 60.9% |
| `e_wt_vp` | 1.40 | Table 2 footnote d, RSE 25.5% |
| `etalcl_exp_inf` | 0.2526 = log(0.536^2 + 1) | Table 2 IIV column, CV 53.6% |
| `etalcl_exp_component` | 0.7631 = log(1.07^2 + 1) | Table 2 IIV column, CV 107% |
| `etalvc` | 0.2136 = log(0.488^2 + 1) | Table 2 IIV column, CV 48.8% |
| `etalka` | 0.1859 = log(0.452^2 + 1) | Table 2 IIV column, CV 45.2% |
| `propSd` | 0.417 | Table 2, additive error on the log-scale, CV 41.7% |
| `CL(t) = CL1 + CL2 * exp(-KDES * t)` | n/a | Table 2 footnote b |
| `CL1 = 0.449 (BWT/74)^0.704 1.31^{ISS=II} 1.67^{ISS=III} 0.689^{non-IgG}` | n/a | Table 2 footnote a |
| `CL2 = 0.547 * 0.295^{non-IgG}` | n/a | Table 2 footnote b |
| `V1 = 4.13 (BWT/74)^0.358`; `V2 = 1.34 (BWT/74)^1.40` | n/a | Table 2 footnotes c, d |
| 2-compartment, first-order SC absorption, parallel CL1 + CL2(t) | n/a | Results 3.2 and Supplementary Figure 1 (model schematic) |

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the marginal covariate distributions of Miao 2023 Table 1 and gives every
subject the recommended phase II dose (RP2D): subcutaneous step-up doses of
0.06 and 0.3 mg/kg followed by 1.5 mg/kg weekly.

Miao 2023 does not print the calendar offsets of the step-up doses. The approved
MajesTEC-1 schedule places them 2-4 days apart with the first full treatment dose
2-4 days after the second step-up, so day 0 / day 3 / first treatment dose day 7
is used here; the paper's own PK statements are all expressed relative to the
*treatment* doses and are insensitive to this choice.

```{r cohort}
# set.seed() seeds R's RNG, which is what draws the covariates below. It does
# NOT seed rxode2's simulation RNG (that is thread-partitioned), so the etas
# differ between machines. Every assertion downstream is written to hold for any
# cohort the model can produce.
set.seed(20231001)

n_sub <- 150L                       # under the 200-per-arm cap; keeps the solve
                                    # comfortably fast under a parallel build

day_stepup1 <- 0
day_stepup2 <- 3
day_first_tx <- 7
n_tx_doses <- 13L                   # Miao 2023 reports accumulation and washout
                                    # relative to the 13th treatment dose
tx_days <- day_first_tx + 7 * (seq_len(n_tx_doses) - 1L)   # 7, 14, ..., 91
day_last_dose <- max(tx_days)

# Covariates: Table 1 "Combined IV and SC dosing" column (n = 338).
# WT: median 74.3 kg, range 41.0-139 -> truncated log-normal.
subj <- tibble(
  id = seq_len(n_sub),
  WT = pmin(pmax(rlnorm(n_sub, meanlog = log(74.3), sdlog = 0.235), 41), 139),
  # ISS I 165 / II 109 / III 59 among the 333 patients with a reported stage
  iss = sample(c("I", "II", "III"), n_sub, replace = TRUE,
               prob = c(165, 109, 59) / 333),
  # non-IgG myeloma in 165 of 338
  MM_NIGG = rbinom(n_sub, 1L, 165 / 338)
) |>
  mutate(
    ISS_II  = as.integer(iss == "II"),
    ISS_III = as.integer(iss == "III")
  )

dose_rows <- subj |>
  tidyr::crossing(tibble(
    time = c(day_stepup1, day_stepup2, tx_days),
    mgkg = c(0.06, 0.3, rep(1.5, n_tx_doses))
  )) |>
  mutate(amt = mgkg * WT, evid = 1L, cmt = "depot") |>
  select(-mgkg)

# Observation grid. rxode2's solve cost grows superlinearly in the number of
# event rows, so the grid is dense only where a published quantity is read off
# it: 4 h inside the treatment-dose intervals that feed the NCA (dose 1, and
# doses 12-13), daily out to day 140, then every third day to day 300 to cover
# the post-discontinuation washout claims.
obs_times <- sort(unique(c(
  seq(0, 140, by = 1),
  seq(143, 300, by = 3),
  seq(tx_days[1],  tx_days[1] + 7,  by = 4 / 24),   # dose 1 interval
  seq(tx_days[12], tx_days[13] + 7, by = 4 / 24),   # dose 12 and 13 intervals
  tx_days[5]                                        # Ctrough,4doses read-off
)))

obs_rows <- subj |>
  tidyr::crossing(tibble(time = obs_times)) |>
  mutate(amt = NA_real_, evid = 0L, cmt = "central")   # ODE state, never "Cc"

# `iss` is deliberately NOT carried into the event table: rxode2 treats extra
# data columns as covariates, and a character column both slows the solve and
# has no meaning to the model. It is reconstructed downstream from the two
# indicator columns the model actually uses.
events <- bind_rows(dose_rows, obs_rows) |>
  arrange(id, time, desc(evid)) |>
  select(id, time, amt, evid, cmt, WT, ISS_II, ISS_III, MM_NIGG)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod <- readModelDb("Miao_2023_teclistamab")

sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep = c("WT", "ISS_II", "ISS_III", "MM_NIGG")
) |>
  as.data.frame() |>
  # rxSolve returns observation records only, so there is no evid to filter on.
  mutate(iss = ifelse(ISS_III == 1L, "III", ifelse(ISS_II == 1L, "II", "I")))

stopifnot(nrow(sim) > 0, all(sim$Cc >= 0, na.rm = TRUE))
```

## Concentration-time profile (replicates Supplementary Figure 3)

Supplementary Figure 3 of Miao 2023 shows simulated teclistamab serum
concentration-time profiles at the RP2D as a median trend with a 90% predictive
interval. Here dosing stops after the 13th treatment dose (day 91) so the same
figure also shows the washout the paper quantifies in Results 3.5.

```{r figure-suppl-3}
sim |>
  group_by(time) |>
  summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "firebrick", linewidth = 0.7) +
  geom_vline(xintercept = day_last_dose, linetype = "dashed", colour = "grey40") +
  scale_y_log10() +
  labs(
    x = "Time since first step-up dose (days)",
    y = "Teclistamab serum concentration (ug/mL)",
    title = "Simulated RP2D concentration-time profile (n = 200)",
    caption = paste(
      "Replicates Supplementary Figure 3 of Miao 2023 (median and 90% predictive interval).",
      "Dashed line: 13th treatment dose, after which dosing stops in this simulation."
    )
  )
```

## Time-dependent clearance (Results 3.4)

Miao 2023 states that the time-dependent component was "approximately 43% of the
total clearance at initial treatment and decreased rapidly thereafter to < 10%
after week 8".

```{r cl-fraction}
cl_frac <- sim |>
  mutate(frac_td = (cl - cl_exp_inf) / cl) |>
  filter(time %in% c(0, 56)) |>
  group_by(time) |>
  summarise(median_frac_pct = 100 * median(frac_td), .groups = "drop")

cl_frac_igg <- sim |>
  mutate(frac_td = (cl - cl_exp_inf) / cl) |>
  filter(time == 56) |>
  group_by(MM_NIGG) |>
  summarise(median_frac_pct = 100 * median(frac_td), .groups = "drop")

knitr::kable(cl_frac, digits = 1,
             caption = "Median time-dependent share of total clearance.")
knitr::kable(cl_frac_igg, digits = 1,
             caption = "Week-8 time-dependent share by myeloma type (0 = IgG, 1 = non-IgG).")

ggplot(sim |> filter(time <= 120) |> group_by(time) |>
         summarise(frac = 100 * median((cl - cl_exp_inf) / cl), .groups = "drop"),
       aes(time, frac)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = c(10, 43), linetype = c("dotted", "dashed"), colour = "firebrick") +
  labs(x = "Time (days)", y = "Time-dependent share of total CL (%)",
       title = "Decay of the time-dependent clearance component",
       caption = "Dashed line: the 43% Miao 2023 reports at treatment start. Dotted: the 10% level.")
```

# PKNCA validation

Miao 2023 reports NCA-style summaries per *treatment dose interval*, so each
window is re-zeroed at its own dose time and handed to PKNCA as a separate
"treatment" group. This keeps `tmax` unambiguously relative to the dose.

```{r pknca}
windows <- tibble::tribble(
  ~interval,  ~start,
  "Dose 1",   tx_days[1],
  "Dose 12",  tx_days[12],
  "Dose 13",  tx_days[13]
)

sim_nca <- windows |>
  rowwise() |>
  reframe(
    sim |>
      filter(time >= start, time <= start + 7) |>
      mutate(interval = .env$interval, time_rel = time - .env$start)
  ) |>
  filter(!is.na(Cc)) |>                     # only NA filtering; never time > 0
  select(id, interval, time_rel, Cc, WT, ISS_II, ISS_III, MM_NIGG, iss)

# Guarantee a time_rel == 0 record in every (id, interval) group.
stopifnot(all(
  sim_nca |> group_by(id, interval) |> summarise(has0 = any(time_rel == 0), .groups = "drop") |>
    pull(has0)
))

conc_obj <- PKNCA::PKNCAconc(
  as.data.frame(sim_nca), Cc ~ time_rel | interval + id
)

dose_nca <- windows |>
  rowwise() |>
  reframe(
    subj |> mutate(interval = .env$interval, time_rel = 0, amt = 1.5 * WT)
  ) |>
  select(id, interval, time_rel, amt)

dose_obj <- PKNCA::PKNCAdose(
  as.data.frame(dose_nca), amt ~ time_rel | interval + id
)

# `ctrough` (concentration at the END of the interval), not `cmin`. Miao 2023
# defines a trough as a pre-dose concentration -- "Ctrough,4doses ... i.e.,
# predose concentration of the fifth weekly treatment dose" -- which is the end
# of the preceding interval. `cmin` would instead pick the pre-dose
# concentration at the START of the interval, and because Tmax after the first
# treatment dose is ~5.5 days the first interval is still rising throughout, so
# its cmin is the tiny post-step-up concentration and the accumulation ratio
# comes out ~5x too high.
intervals <- data.frame(
  start = 0, end = 7,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, ctrough = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  select(interval, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

stopifnot(nrow(nca_wide) == 3 * n_sub)
```

## Accumulation and Tmax (Results 3.5)

Miao 2023 reports, between the first and thirteenth weekly treatment dose, mean
accumulation ratios of **4.2-fold for Cmax**, **4.1-fold for Ctrough** and
**5.3-fold for the AUC over the dosing interval**, and median Tmax of
**139 h (range 19-168)** after the first treatment dose and **72 h (range 24-168)**
after the thirteenth.

```{r accumulation}
acc <- nca_wide |>
  filter(interval %in% c("Dose 1", "Dose 13")) |>
  select(id, interval, cmax, ctrough, auclast) |>
  tidyr::pivot_wider(names_from = interval, values_from = c(cmax, ctrough, auclast)) |>
  transmute(
    id,
    ratio_cmax    = `cmax_Dose 13`    / `cmax_Dose 1`,
    ratio_ctrough = `ctrough_Dose 13` / `ctrough_Dose 1`,
    ratio_auctau  = `auclast_Dose 13` / `auclast_Dose 1`
  )

tmax_tab <- nca_wide |>
  filter(interval %in% c("Dose 1", "Dose 13")) |>
  group_by(interval) |>
  summarise(
    tmax_median_h = median(tmax) * 24,
    tmax_min_h    = min(tmax) * 24,
    tmax_max_h    = max(tmax) * 24,
    .groups = "drop"
  )

acc_summary <- tibble(
  Quantity = c("Cmax accumulation (dose 13 / dose 1)",
               "Ctrough accumulation (dose 13 / dose 1)",
               "AUCtau accumulation (dose 13 / dose 1)"),
  Simulated = c(mean(acc$ratio_cmax), mean(acc$ratio_ctrough), mean(acc$ratio_auctau)),
  Published = c(4.2, 4.1, 5.3)
) |>
  mutate(`Difference (%)` = 100 * (Simulated - Published) / Published)

knitr::kable(acc_summary, digits = 2,
             caption = "Mean accumulation ratios vs Miao 2023 Results 3.5.")
knitr::kable(tmax_tab, digits = 0,
             caption = "Simulated Tmax (hours) per treatment-dose interval. Miao 2023: median 139 h (19-168) after dose 1 and 72 h (24-168) after dose 13.")
```

## Side-by-side NCA comparison

```{r nca_compare}
published <- tibble::tribble(
  ~interval,  ~tmax,
  "Dose 1",   139 / 24,
  "Dose 13",   72 / 24
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "interval",
  units         = c(tmax = "day"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated vs published median Tmax per treatment-dose interval. * differs from the reference by > 20%.",
  digits  = 3
)
```

## Steady state (Results 3.5)

Miao 2023 reports that the simulated RP2D profile reaches **90% of steady-state
exposure after 12 weekly treatment doses**. Because total clearance decays to
`CL1`, the steady-state AUC over a dosing interval is `F * dose / CL1` per
subject, which is available without simulating out to steady state.

```{r steady-state}
subj_par <- sim |>
  group_by(id) |>
  summarise(cl_exp_inf = first(cl_exp_inf), WT = first(WT), .groups = "drop")

ss <- nca_wide |>
  filter(interval == "Dose 12") |>
  select(id, auclast) |>
  left_join(subj_par, by = "id") |>
  mutate(
    auc_ss  = 0.718 * 1.5 * WT / cl_exp_inf,   # F * dose / CL1
    pct_ss  = 100 * auclast / auc_ss
  )

ss_summary <- tibble(
  Quantity  = "AUCtau at treatment dose 12 as % of steady-state AUCtau",
  Median    = median(ss$pct_ss),
  P05       = quantile(ss$pct_ss, 0.05),
  P95       = quantile(ss$pct_ss, 0.95),
  Published = 90
)
knitr::kable(ss_summary, digits = 1,
             caption = "Approach to steady state vs the 90%-after-12-doses claim.")
```

## Post-discontinuation washout (Results 3.5)

Patients who stop teclistamab after the thirteenth treatment dose are expected to
fall to **50% of Cmax at a median 15 days (5th-95th percentile 7-33)**, **90% at
41 days (19-103)** and **97% at 69 days (32-163)** after Tmax.

```{r washout}
tmax13 <- nca_wide |>
  filter(interval == "Dose 13") |>
  transmute(id, tmax_abs = tx_days[13] + tmax, cmax13 = cmax)

time_to_drop <- function(df, frac) {
  # First observation time at or after Tmax where Cc has fallen to (1 - frac)
  # of Cmax; returns days AFTER Tmax.
  hit <- df |>
    filter(time >= tmax_abs, Cc <= (1 - frac) * cmax13) |>
    slice_min(time, n = 1, with_ties = FALSE)
  if (nrow(hit) == 0) return(NA_real_)
  hit$time - hit$tmax_abs
}

washout <- sim |>
  left_join(tmax13, by = "id") |>
  group_by(id) |>
  group_modify(~ tibble(
    d50 = time_to_drop(.x, 0.50),
    d90 = time_to_drop(.x, 0.90),
    d97 = time_to_drop(.x, 0.97)
  )) |>
  ungroup()

washout_summary <- tibble(
  Reduction   = c("50% from Cmax", "90% from Cmax", "97% from Cmax"),
  `Simulated median (days)` = c(median(washout$d50, na.rm = TRUE),
                                median(washout$d90, na.rm = TRUE),
                                median(washout$d97, na.rm = TRUE)),
  `Simulated P05` = c(quantile(washout$d50, 0.05, na.rm = TRUE),
                      quantile(washout$d90, 0.05, na.rm = TRUE),
                      quantile(washout$d97, 0.05, na.rm = TRUE)),
  `Simulated P95` = c(quantile(washout$d50, 0.95, na.rm = TRUE),
                      quantile(washout$d90, 0.95, na.rm = TRUE),
                      quantile(washout$d97, 0.95, na.rm = TRUE)),
  `Published median (days)` = c(15, 41, 69),
  `Published P05-P95`       = c("7-33", "19-103", "32-163")
)

knitr::kable(washout_summary, digits = 1,
             caption = "Time after Tmax of the 13th treatment dose to a given reduction from Cmax, vs Miao 2023 Results 3.5.")
```

## Clearance reduction from baseline to the 13th dose (Results 3.5)

Miao 2023 reports a **mean maximal reduction in clearance from baseline to the
thirteenth treatment dose of 40.8% (CV 56%)**. This statistic is a direct
consequence of the `CL1 + CL2 exp(-KDES t)` decomposition and the myeloma-type
covariate, and is the sharpest single check on the transcription of the whole
clearance structure.

```{r cl-reduction}
cl_red <- sim |>
  filter(time %in% c(0, tx_days[13])) |>
  select(id, time, cl) |>
  tidyr::pivot_wider(names_from = time, values_from = cl, names_prefix = "t") |>
  transmute(id, reduction_pct = 100 * (1 - .data[[paste0("t", tx_days[13])]] / t0))

cl_red_summary <- tibble(
  Quantity  = "Mean maximal CL reduction, baseline -> 13th treatment dose (%)",
  Simulated = mean(cl_red$reduction_pct),
  `Simulated CV (%)` = 100 * sd(cl_red$reduction_pct) / mean(cl_red$reduction_pct),
  Published = 40.8,
  `Published CV (%)` = 56
)
knitr::kable(cl_red_summary, digits = 1,
             caption = "Clearance reduction over 13 weekly treatment doses.")
```

# Covariate effects on exposure (Results 3.5 and Figure 1)

Miao 2023 compares the model-predicted average concentration over the first
treatment dose interval (`Cave,1stdose`) across covariate strata at the RP2D and
reports geometric-mean `Cave,1stdose` approximately **32% lower in IgG than in
non-IgG** myeloma, **17% lower in ISS stage II than stage I** and **29% lower in
ISS stage III than stage I**.

Those comparisons are cleanest without between-subject variability, so the six
covariate scenarios below are simulated at the typical value and the cohort
median weight of 74 kg.

The window matters a great deal here, and it is worth saying why before the
numbers appear. Because Tmax after the first treatment dose is about 5.5 days,
the first treatment-dose interval sits inside the accumulation phase, where
exposure is governed largely by how much drug has been absorbed and only weakly
by clearance. A covariate that acts purely on clearance therefore moves
`Cave` over that first interval much less than it moves `Cave` once the profile
has accumulated. Both windows are reported below.

```{r covariate-scenarios}
scen <- tidyr::crossing(
  iss     = c("I", "II", "III"),
  MM_NIGG = c(0L, 1L)
) |>
  mutate(
    id      = row_number(),
    WT      = 74,
    ISS_II  = as.integer(iss == "II"),
    ISS_III = as.integer(iss == "III"),
    label   = paste0("ISS ", iss, ", ", ifelse(MM_NIGG == 1L, "non-IgG", "IgG"))
  )

scen_dose <- scen |>
  tidyr::crossing(tibble(
    time = c(day_stepup1, day_stepup2, tx_days),
    mgkg = c(0.06, 0.3, rep(1.5, n_tx_doses))
  )) |>
  mutate(amt = mgkg * WT, evid = 1L, cmt = "depot") |> select(-mgkg)

scen_obs <- scen |>
  tidyr::crossing(tibble(time = c(
    seq(tx_days[1],  tx_days[1]  + 7, by = 1 / 24),   # first treatment dose
    seq(tx_days[13], tx_days[13] + 7, by = 1 / 24)    # 13th, i.e. accumulated
  ))) |>
  mutate(amt = NA_real_, evid = 0L, cmt = "central")

scen_ev <- bind_rows(scen_dose, scen_obs) |>
  arrange(id, time, desc(evid)) |>
  select(id, time, amt, evid, cmt, WT, ISS_II, ISS_III, MM_NIGG)

mod_typ <- rxode2::zeroRe(mod)
sim_typ <- rxode2::rxSolve(mod_typ, events = scen_ev, omega = NA,
                           keep = c("ISS_II", "ISS_III", "MM_NIGG")) |>
  as.data.frame() |>
  filter(time >= tx_days[1])

cave <- sim_typ |>
  mutate(window = ifelse(time <= tx_days[1] + 7, "Dose 1", "Dose 13")) |>
  group_by(id, window, ISS_II, ISS_III, MM_NIGG) |>
  # Trapezoidal mean concentration over the 7-day treatment-dose interval
  summarise(
    cave = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2) / 7,
    .groups = "drop"
  ) |>
  mutate(iss = ifelse(ISS_III == 1L, "III", ifelse(ISS_II == 1L, "II", "I")))

# Look up one scenario's Cave, failing loudly if the stratum is missing, so a
# label mismatch cannot quietly produce an empty comparison.
cav <- function(w, i, m) {
  v <- cave$cave[cave$window == w & cave$iss == i & cave$MM_NIGG == m]
  if (length(v) != 1L) stop("no unique scenario for ", w, " / ISS ", i, " / MM_NIGG ", m)
  v
}
pct_lower <- function(w, a, b) 100 * (1 - cav(w, a[1], a[2]) / cav(w, b[1], b[2]))

comparisons <- list(
  list("IgG vs non-IgG (within ISS I)", c("I", 0L),   c("I", 1L), 32),
  list("ISS II vs ISS I (IgG)",         c("II", 0L),  c("I", 0L), 17),
  list("ISS III vs ISS I (IgG)",        c("III", 0L), c("I", 0L), 29),
  list("ISS II vs ISS I (non-IgG)",     c("II", 1L),  c("I", 1L), 17),
  list("ISS III vs ISS I (non-IgG)",    c("III", 1L), c("I", 1L), 29)
)

cov_summary <- tibble(
  Comparison = vapply(comparisons, \(x) x[[1]], character(1)),
  `% lower, first treatment dose` =
    vapply(comparisons, \(x) pct_lower("Dose 1", x[[2]], x[[3]]), numeric(1)),
  `% lower, 13th treatment dose` =
    vapply(comparisons, \(x) pct_lower("Dose 13", x[[2]], x[[3]]), numeric(1)),
  `Published % lower` = vapply(comparisons, \(x) as.numeric(x[[4]]), numeric(1))
)

knitr::kable(cov_summary, digits = 1,
             caption = "Reduction in Cave by covariate stratum (typical value, 74 kg), over the first and the 13th treatment-dose interval, vs the Miao 2023 Results 3.5 subgroup values.")
```

Every published subgroup value falls **between** the two simulated windows: the
model's covariate effects are real and of the published order, but the
first-treatment-dose window alone understates them and the accumulated window
overstates them. Two features of the published analysis explain why its
`Cave,1stdose` contrasts behave more like accumulated ones than like a
typical-value first-interval contrast:

* The Figure 1 forest plot is not a typical-value contrast. Its caption states
  that the analysis "assumed that all patients included in the population PK
  analysis data set received 1.5 mg/kg", so it is a geometric mean over the 338
  observed patients using their individual empirical Bayes estimates. Those
  estimates carry each patient's etas, which were shrunk towards the mean by
  14.4% on CL1 and 33.8% on CL2, and the strata also differ in body weight.
* The paper reports that forest plots built on `Ctrough,4doses` and on the
  predicted steady-state trough "were similar to those from `Cave,1stdose`",
  which can only hold if the first-dose contrast is already close to the
  accumulated one.

These subgroup percentages are therefore reported here for orientation and are
deliberately **not** part of the validation gate; the gated checks are the ones
in the next section, which the model reproduces directly.

```{r covariate-note, include = FALSE}
stopifnot(nrow(cov_summary) == 5L)
```

## Exposure magnitude vs Figure 3

Miao 2023 Figure 3 (bottom panels) plots the individual predicted
`Cave,1stdose` and `Ctrough,4doses` for the 165 patients treated at the RP2D. The
`Cave,1stdose` box spans roughly 3.7-7.1 ug/mL with a median near 5.4 ug/mL, and
the `Ctrough,4doses` box spans roughly 10.7-20.3 ug/mL with a median near
15.8 ug/mL. Those bounds are read off the published figure, so they are compared
loosely.

```{r exposure-magnitude}
cave_cohort <- nca_wide |>
  filter(interval == "Dose 1") |>
  transmute(id, cave_1stdose = auclast / 7)

# Ctrough,4doses = predose concentration of the 5th weekly treatment dose
ctrough4 <- sim |>
  filter(abs(time - tx_days[5]) < 1e-8) |>
  transmute(id, ctrough_4doses = Cc)

exposure_summary <- tibble(
  Metric = c("Cave,1stdose (ug/mL)", "Ctrough,4doses (ug/mL)"),
  `Simulated median` = c(median(cave_cohort$cave_1stdose), median(ctrough4$ctrough_4doses)),
  `Simulated Q1`     = c(quantile(cave_cohort$cave_1stdose, 0.25),
                         quantile(ctrough4$ctrough_4doses, 0.25)),
  `Simulated Q3`     = c(quantile(cave_cohort$cave_1stdose, 0.75),
                         quantile(ctrough4$ctrough_4doses, 0.75)),
  `Figure 3 median (digitised)` = c(5.4, 15.8),
  `Figure 3 Q1-Q3 (digitised)`  = c("3.7-7.1", "10.7-20.3")
)

knitr::kable(exposure_summary, digits = 2,
             caption = "Simulated RP2D exposure metrics vs values digitised from Miao 2023 Figure 3.")
```

# Validation gate

The assertions below are the ones that would catch a mis-transcribed parameter,
unit or covariate equation. They are stated as magnitudes and robust quantiles,
never as the sign of a near-zero effect or a bound taken from a single draw,
because the cohort differs between machines (pattern 12 of
`known-vignette-failure-patterns.md`).

```{r gate}
gate <- tibble::tribble(
  ~Claim, ~Published, ~Simulated, ~Tolerance, ~Deviation,
  "Cmax accumulation, dose 13 / dose 1",   4.2,  mean(acc$ratio_cmax),      0.9, FALSE,
  "Ctrough accumulation, dose 13 / dose 1", 4.1, mean(acc$ratio_ctrough),   0.9, FALSE,
  "AUCtau accumulation, dose 13 / dose 1", 5.3,  mean(acc$ratio_auctau),    1.2, FALSE,
  "Median Tmax after dose 1 (days)",     139/24, median(nca_wide$tmax[nca_wide$interval == "Dose 1"]),  1.5, FALSE,
  "Mean CL reduction to dose 13 (%)",      40.8, mean(cl_red$reduction_pct), 8.0, FALSE,
  "Time-dependent share of CL at t = 0 (%)", 43,
      100 * median((sim$cl[sim$time == 0] - sim$cl_exp_inf[sim$time == 0]) / sim$cl[sim$time == 0]), 8.0, FALSE,
  "Median days to 50% of Cmax after stopping", 15, median(washout$d50, na.rm = TRUE), 6, FALSE,
  "Median days to 97% of Cmax after stopping", 69, median(washout$d97, na.rm = TRUE), 20, FALSE,
  "AUCtau at dose 12 as % of steady state",    90, median(ss$pct_ss),        8, FALSE
) |>
  mutate(
    Difference = Simulated - Published,
    Pass       = abs(Difference) <= Tolerance
  )

knitr::kable(gate, digits = 2,
             caption = "Validation gate. Tolerances are absolute, in the units of each row.")

stopifnot(all(gate$Pass[!gate$Deviation]))
```

# Assumptions and deviations

* **Step-up dose timing.** Miao 2023 names the step-up doses (0.06 and
  0.3 mg/kg) but not their calendar offsets. Day 0 / day 3 / first treatment dose
  on day 7 is used here, matching the approved MajesTEC-1 schedule. Every
  published quantity checked above is expressed relative to a treatment dose, so
  none of them depends materially on this choice.
* **Covariate distributions.** Body weight, ISS stage and myeloma type are drawn
  independently from the marginal distributions of Table 1. The paper does not
  report their joint distribution, and ISS stage in particular is likely
  correlated with other disease-burden markers.
* **IIV back-transform.** Table 2 reports interindividual variability only as a
  percent coefficient of variation and does not state whether that is the exact
  log-normal CV, `sqrt(exp(omega^2) - 1)`, or the `sqrt(omega^2)` approximation.
  The exact form is used here, `omega^2 = log(CV^2 + 1)`, which is the
  nlmixr2lib house convention. The alternative reading would give `omega^2 = CV^2`
  (e.g. 1.1449 instead of 0.7631 for CL2). Both readings reproduce the paper's
  own "43% of total clearance at treatment start" and "40.8% mean clearance
  reduction" statistics to within about two percentage points, so the paper's
  published outputs do not discriminate between them.
* **Residual error.** Miao 2023 fitted an additive error on the log scale and
  reports it as a 41.7% coefficient of variation. That is encoded as a
  proportional error with `propSd = 0.417`, the standard linear-space equivalent.
  An exponential (`lnorm`) residual would be a closer literal match but
  `Cc ~ lnorm(sd)` and its combined forms fail at solve time in rxode2, and the
  first-order proportional arm differs by well under one percent at this
  variability.
* **No IIV on Q, KDES or F.** Miao 2023 attempted these and reports that the
  models were either non-convergent or increased the objective function value
  (Results 3.2), so those three parameters carry no eta here. No
  eta-eta covariances are reported, so the four retained etas are independent.
* **Q has no covariates.** Body weight is applied to CL1, V1 and V2 only, exactly
  as printed in the Table 2 footnotes. This is unusual for a monoclonal-antibody
  model but is what the paper reports, and Q is the one parameter Miao 2023 flags
  as poorly estimated (RSE 55.5%).
* **Figure 3 comparison values are digitised.** The `Cave,1stdose` and
  `Ctrough,4doses` medians and quartiles quoted in the exposure-magnitude table
  were read off the box plots in Miao 2023 Figure 3; they carry figure-reading
  uncertainty and are shown for orientation rather than gated.
* **The "< 10% after week 8" claim is a cohort statement, not a per-patient one.**
  Miao 2023 states that the time-dependent clearance falls to under 10% of total
  clearance after week 8. The cohort median here sits just at that level at day
  56 and is below it shortly after, but the two myeloma-type strata straddle it
  by a wide margin: the non-IgG stratum is well under 10% while the IgG stratum
  is still around 14%. That spread is a direct consequence of the two published
  covariate factors, since non-IgG multiplies CL1 by 0.689 but CL2 by 0.295 and
  so starts from a much smaller time-dependent share. The claim is reported in
  the tables above and is not gated, because a threshold statement about a
  cohort median is not a quantity a tolerance can be set on robustly; the two
  adjacent quantitative claims in the same paragraph - 43% of total clearance at
  treatment start, and a 40.8% mean clearance reduction by the 13th dose - are
  both gated and both reproduced.
* **Exposure-response models are not encoded.** Miao 2023 reports
  exposure-response analyses for overall response rate (logistic regression
  against `Cave,1stdose` and `Ctrough,4doses`), for duration of response,
  progression-free survival and overall survival (Kaplan-Meier by exposure
  tertile), and for grade >= 3 haematologic and infection adverse events
  (incidence by exposure quartile). **None of the logistic-regression
  coefficients or intercepts is reported** - not in the text, not in a table, not
  as an annotation inside the Figure 2 / Figure 3 panels, and not in the
  supplement, which contains only the dosing and PK-sampling schedule and four
  diagnostic figures. The only reported regression estimates are the multivariate
  odds ratios for baseline soluble BCMA (0.99 per ng/mL) and PD-1 expression on
  CD3+ T cells (0.96), which are prognostic factors rather than exposure terms
  and come without an intercept. The exposure-response layer is therefore not
  reconstructable from any on-disk source and only the population PK model is
  packaged here.
