---
title: "Riociguat (Michalickova 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Riociguat (Michalickova 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Michalickova_2020_riociguat"))$reference`
* Description: `r rxode2::rxode(readModelDb("Michalickova_2020_riociguat"))$description`
* Article: <https://doi.org/10.1177/2045894019898031> (open access)

Riociguat is a soluble guanylate cyclase stimulator used for chronic
thromboembolic pulmonary hypertension (CTEPH). Its main circulating metabolite,
M1 (desmethylriociguat), is pharmacologically active. Michalickova 2020 fitted a
joint parent + metabolite model to sparse therapeutic-drug-monitoring data from
routine clinical practice. Each patient contributed one riociguat and one M1
serum concentration.

# Population

Forty-nine adults with CTEPH (24 female, 25 male) were treated at the General
University Hospital in Prague, Czech Republic. Thirty-seven (74%) had inoperable
CTEPH and 13 (26%) had persistent or recurrent pulmonary hypertension after
pulmonary endarterectomy. Median (IQR) age was 74 (66-78) years, body weight 80
(67-95) kg, creatinine clearance (CKD-EPI) 70 (59-79) mL/min and total bilirubin
0.69 (0.53-0.98) mg/dL (Michalickova 2020 Table 1). Everyone had been on a stable
riociguat dose of 1.5-2.5 mg three times daily for at least three months (median
7.5 mg/day, IQR 6.75-7.5 mg/day). The single steady-state sample was drawn 1.25
to 6.75 h after the last dose. Riociguat concentrations ranged from 44 to 749
ug/L and M1 concentrations from 17 to 314 ug/L.

The same information is available programmatically via
`readModelDb("Michalickova_2020_riociguat")()$population`.

# Source trace

The in-file comments next to each `ini()` entry in
`inst/modeldb/specificDrugs/Michalickova_2020_riociguat.R` record where each
value came from. This table collects them in one place.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` (fixed) | log(3) 1/h | Table 2 'Ka/F 3 FIX'; Methods (literature value, ref 17) |
| `lcl_nonmet` (CLe,r/F at CRCL 70 mL/min) | log(0.66) L/h | Table 2 |
| `lcl_met` (CLf,M1/F at bilirubin 0.69 mg/dL) | log(0.665) L/h | Table 2 |
| `e_tbili_cl_met` | -0.462 | Table 2 (Results text: -0.463) |
| `lvc` (VP/F = VM/F) | log(3.63) L | Table 2; Methods (volumes assumed equal) |
| `lcl_m1` (CLe,M1/F) | log(1.47) L/h | Table 2 |
| `ltlag` (six outlying patients) | log(2.95) h | Table 2 'Tlag (ID = 6,10,11,19,25,28)' |
| `propSd` | sqrt(0.152) = 0.390 | Table 2, riociguat proportional variance |
| `propSd_m1` | sqrt(0.268) = 0.518 | Table 2, M1 proportional variance |
| CLe,r/F = CLe,rTV * (CREACL / 70) | n/a | Table 2; Results ('increase 0.009 L/h per unit') |
| CLf,M1/F = CLf,M1TV * (BILTOT / 0.69)^theta | n/a | Table 2; Results |
| One-compartment parent -> one-compartment M1, parallel CLe,r | n/a | Figure 2; Results |
| Lag time only when MIX_LAGGED_ABS = 1 | n/a | Methods 'Covariate analysis'; Results |

# Typical-value steady-state profiles

The patients were at steady state on three-times-daily dosing, so every
simulation below starts from a steady-state dose (`ss = 1`, `ii = 8`). The model
has two endpoints (`Cc` and `Cc_m1`) and neither is an ODE state. Observation
rows are therefore keyed by `dvid` (1 = riociguat, 2 = M1) with `cmt` left
empty. Typical-value solves use `omega = NA, sigma = NA`.

```{r typical-events}
mod <- readModelDb("Michalickova_2020_riociguat")

ref_cov <- tibble(CRCL = 70, TBILI = 0.69 * 17.1) # cohort medians; TBILI in umol/L

make_ss_events <- function(dose_mg, lagged, id, obs_times = seq(0, 8, by = 0.1)) {
  dose <- tibble(
    id = id, time = 0, amt = dose_mg, evid = 1L, cmt = "depot",
    ii = 8, ss = 1L, dvid = NA_integer_
  )
  obs <- tidyr::expand_grid(time = obs_times, dvid = 1:2) |>
    mutate(
      id = id, amt = NA_real_, evid = 0L, cmt = NA_character_,
      ii = 0, ss = 0L
    )
  bind_rows(dose, obs) |>
    mutate(
      MIX_LAGGED_ABS = lagged, dose_mg = dose_mg,
      CRCL = ref_cov$CRCL, TBILI = ref_cov$TBILI
    )
}

scen <- tidyr::expand_grid(dose_mg = c(1.5, 2, 2.5), lagged = c(0, 1)) |>
  mutate(id = seq_len(n()))
ev_typ <- lapply(seq_len(nrow(scen)), function(i) {
  make_ss_events(scen$dose_mg[i], scen$lagged[i], scen$id[i])
}) |>
  bind_rows() |>
  arrange(id, time, desc(evid))

sim_typ <- rxode2::rxSolve(
  mod, ev_typ,
  omega = NA, sigma = NA, useLinCmt = FALSE,
  returnType = "data.frame",
  keep = c("dose_mg", "MIX_LAGGED_ABS"),
  rtol = 1e-10, atol = 1e-12, ssRtol = 1e-10, ssAtol = 1e-12
) |>
  distinct(id, time, .keep_all = TRUE) |>
  mutate(
    regimen = paste0(dose_mg, " mg TID"),
    absorption = ifelse(MIX_LAGGED_ABS == 1, "2.95 h lag", "no lag")
  )
```

```{r typical-profiles, fig.width = 7, fig.height = 4}
sim_typ |>
  select(time, regimen, absorption, riociguat = Cc, M1 = Cc_m1) |>
  pivot_longer(c(riociguat, M1), names_to = "analyte", values_to = "conc") |>
  ggplot(aes(time, conc, colour = regimen, linetype = absorption)) +
  geom_line() +
  facet_wrap(~analyte, scales = "free_y") +
  labs(
    x = "Time after dose at steady state (h)",
    y = "Serum concentration (ug/L)",
    title = "Typical-patient steady-state profiles",
    caption = "CRCL 70 mL/min, total bilirubin 0.69 mg/dL."
  )
```

## Closed-form check

For a linear model the steady-state average concentration over a dosing
interval follows directly from the clearances. Riociguat averages
`Dose / tau / (CLf,M1/F + CLe,r/F)`. M1 averages
`Dose / tau * CLf,M1 / (CLf,M1 + CLe,r) / CLe,M1`. The lag time shifts the
profile in time and leaves both averages unchanged. PKNCA's `cav` over one
steady-state interval must reproduce these closed forms. The two sides use
identical parameters, so the tolerance only has to cover numerical integration
error.

```{r closed-form}
cl_met <- 0.665
cl_nonmet <- 0.66
cl_m1 <- 1.47
closed_form <- scen |>
  mutate(
    regimen = paste0(dose_mg, " mg TID"),
    absorption = ifelse(lagged == 1, "2.95 h lag", "no lag"),
    cav_parent = dose_mg / 8 / (cl_met + cl_nonmet) * 1000,
    cav_m1 = dose_mg / 8 * cl_met / (cl_met + cl_nonmet) / cl_m1 * 1000
  )

run_nca <- function(sim, conc_col) {
  conc <- sim |>
    filter(!is.na(.data[[conc_col]])) |>
    transmute(id, time, conc = .data[[conc_col]], regimen, absorption)
  dose <- sim |>
    distinct(id, regimen, absorption, dose_mg) |>
    mutate(time = 0)
  o_conc <- PKNCA::PKNCAconc(conc, conc ~ time | regimen + absorption + id)
  o_dose <- PKNCA::PKNCAdose(dose, dose_mg ~ time | regimen + absorption + id)
  intervals <- data.frame(
    start = 0, end = 8, cmax = TRUE, tmax = TRUE, cmin = TRUE,
    auclast = TRUE, cav = TRUE
  )
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals))
  as.data.frame(res)
}

nca_parent <- run_nca(sim_typ, "Cc")
nca_m1 <- run_nca(sim_typ, "Cc_m1")

cav_check <- bind_rows(
  nca_parent |> filter(PPTESTCD == "cav") |> mutate(analyte = "riociguat"),
  nca_m1 |> filter(PPTESTCD == "cav") |> mutate(analyte = "M1")
) |>
  left_join(closed_form, by = c("regimen", "absorption")) |>
  mutate(
    closed = ifelse(analyte == "riociguat", cav_parent, cav_m1),
    rel_err = PPORRES / closed - 1
  )

# PKNCA integrates a 0.1 h grid with the linear-up/log-down trapezoid, so the
# difference is trapezoidal error on the narrow absorption peak (~0.1-0.3%
# measured), not model error. A mis-transcribed clearance or volume unit would
# move these averages by tens of percent.
stopifnot(max(abs(cav_check$rel_err)) < 0.01)

cav_check |>
  select(analyte, regimen, absorption, PPORRES, closed, rel_err) |>
  mutate(rel_err = sprintf("%.3f%%", 100 * rel_err)) |>
  rename(
    "Analyte" = analyte, "Regimen" = regimen, "Absorption" = absorption,
    "Cav simulated (ug/L)" = PPORRES, "Cav closed form (ug/L)" = closed,
    "Relative difference" = rel_err
  ) |>
  knitr::kable(digits = 1)
```

## PKNCA steady-state summary

The paper reports no NCA summary. The table shows the model's steady-state
exposures for the typical patient at the three prescribed dose levels, without
the lag time. Both analytes are side by side, with the closed-form `Cav` as the
reference column.

```{r nca-table}
nca_ref <- closed_form |>
  filter(lagged == 0) |>
  transmute(regimen, analyte = "riociguat", cav = cav_parent) |>
  bind_rows(
    closed_form |> filter(lagged == 0) |>
      transmute(regimen, analyte = "M1", cav = cav_m1)
  )
nca_sim <- bind_rows(
  nca_parent |> mutate(analyte = "riociguat"),
  nca_m1 |> mutate(analyte = "M1")
) |>
  filter(absorption == "no lag") |>
  select(regimen, analyte, PPTESTCD, PPORRES)

tab <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_sim,
  reference = nca_ref,
  by = c("regimen", "analyte"),
  params = "cav",
  units = c(cav = "ug/L")
)
knitr::kable(tab)

nca_sim |>
  filter(PPTESTCD %in% c("cmax", "tmax", "cmin")) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  rename(
    "Regimen" = regimen, "Analyte" = analyte, "Cmax,ss (ug/L)" = cmax,
    "Tmax (h)" = tmax, "Cmin,ss (ug/L)" = cmin
  ) |>
  knitr::kable(digits = 2)
```

# Covariate relationships (Figure 3)

```{r figure-3, fig.width = 7, fig.height = 3.5}
# Replicates Figure 3 of Michalickova 2020: typical CLf,M1/F versus total
# bilirubin and CLe,r/F versus creatinine clearance.
bili <- tibble(tbili_mgdl = seq(0.3, 2.0, length.out = 100)) |>
  mutate(value = 0.665 * (tbili_mgdl / 0.69)^-0.462, panel = "CLf,M1/F vs total bilirubin (mg/dL)", x = tbili_mgdl)
crcl <- tibble(CRCL = seq(30, 120, length.out = 100)) |>
  mutate(value = 0.66 * (CRCL / 70), panel = "CLe,r/F vs creatinine clearance (mL/min)", x = CRCL)
bind_rows(bili, crcl) |>
  ggplot(aes(x, value)) +
  geom_line() +
  facet_wrap(~panel, scales = "free_x") +
  labs(
    x = NULL, y = "Apparent clearance (L/h)",
    caption = "Replicates Figure 3 of Michalickova 2020 (typical-value curves)."
  )

# The printed covariate equations must pass through the Table 2 typical
# values at the reference covariates and reproduce the Results slope.
stopifnot(
  abs(0.66 * (70 / 70) - 0.66) < 1e-12,
  abs(0.665 * (0.69 / 0.69)^-0.462 - 0.665) < 1e-12,
  abs(0.66 / 70 - 0.009) < 0.0005 # Results: '0.009 L/h per unit (mL/min)'
)
```

# Virtual cohort and Figure 1

Figure 1 of the paper plots the single observed riociguat and M1 concentration
of each patient against the time since the last dose. The virtual cohort draws
200 patients. Creatinine clearance and total bilirubin are log-normal with the
Table 1 medians and a spread matched to the IQRs. Doses are 2.5 mg TID for 70%
of patients, 2 mg for 20% and 1.5 mg for 10%, which matches the Table 1 median
and IQR. Each patient has one sampling time drawn uniformly over the observed
1.25-6.75 h window. `MIX_LAGGED_ABS` is Bernoulli(6/49). With one sample per
patient the paper could not separate inter-individual from residual
variability, so the proportional error terms carry all of it and the `sim`
column is the full predictive distribution.

```{r cohort}
set.seed(20200101)
n_sub <- 200
cohort <- tibble(
  id = seq_len(n_sub),
  # IQR of a log-normal = median * exp(+/- 0.674 sigma)
  CRCL = 70 * exp(rnorm(n_sub, 0, log(79 / 59) / (2 * 0.674))),
  tbili_mgdl = 0.69 * exp(rnorm(n_sub, 0, log(0.98 / 0.53) / (2 * 0.674))),
  dose_mg = sample(c(2.5, 2, 1.5), n_sub, replace = TRUE, prob = c(0.7, 0.2, 0.1)),
  MIX_LAGGED_ABS = rbinom(n_sub, 1, 6 / 49),
  tobs = runif(n_sub, 1.25, 6.75)
) |>
  mutate(TBILI = tbili_mgdl * 17.1)

ev_cohort <- bind_rows(
  cohort |> transmute(
    id, time = 0, amt = dose_mg, evid = 1L, cmt = "depot", ii = 8, ss = 1L,
    dvid = NA_integer_, CRCL, TBILI, MIX_LAGGED_ABS
  ),
  tidyr::expand_grid(cohort, dvid = 1:2) |> transmute(
    id, time = tobs, amt = NA_real_, evid = 0L, cmt = NA_character_, ii = 0,
    ss = 0L, dvid, CRCL, TBILI, MIX_LAGGED_ABS
  )
) |>
  arrange(id, time, desc(evid), dvid)

sim_vpc <- suppressWarnings(rxode2::rxSolve(
  mod, ev_cohort,
  useLinCmt = FALSE, returnType = "data.frame",
  keep = c("MIX_LAGGED_ABS")
)) |>
  group_by(id) |>
  mutate(analyte = c("riociguat", "M1")[row_number()]) |>
  ungroup()

# Each subject has exactly two observation rows (dvid 1 then 2). Confirm the
# row-to-endpoint mapping from the output itself: on the riociguat row the
# residual-free prediction ipredSim equals Cc, on the M1 row it equals Cc_m1.
stopifnot(
  all(count(sim_vpc, id)$n == 2),
  with(filter(sim_vpc, analyte == "riociguat"), all(abs(ipredSim / Cc - 1) < 1e-8)),
  with(filter(sim_vpc, analyte == "M1"), all(abs(ipredSim / Cc_m1 - 1) < 1e-8))
)
```

```{r figure-1, fig.width = 7, fig.height = 5}
# Replicates Figure 1 of Michalickova 2020 with simulated observations.
observed_range <- tibble(
  analyte = c("riociguat", "M1"), lo = c(44, 17), hi = c(749, 314)
)
sim_vpc |>
  mutate(absorption = ifelse(MIX_LAGGED_ABS == 1, "2.95 h lag", "no lag")) |>
  ggplot(aes(time, sim)) +
  geom_rect(
    data = observed_range, inherit.aes = FALSE,
    aes(xmin = 1.25, xmax = 6.75, ymin = lo, ymax = hi),
    fill = "grey85", alpha = 0.6
  ) +
  geom_point(aes(shape = absorption), alpha = 0.7) +
  facet_wrap(~analyte, ncol = 1, scales = "free_y") +
  labs(
    x = "Time after the last dose (h)", y = "Serum concentration (ug/L)",
    shape = NULL,
    caption = paste(
      "Simulated single steady-state samples (n = 200). Grey box: observed",
      "time window and concentration range reported by Michalickova 2020."
    )
  )
```

The published figure shows riociguat mostly between about 60 and 600 ug/L and
M1 between about 20 and 300 ug/L. The simulated cloud occupies the same region.
The check below is on the centre and on robust quantiles of the simulated
distribution, not on its extremes (see the repository notes on cohort
assertions). It confirms that most simulated samples fall inside the observed
range and that the simulated medians match the observed data.

```{r figure-1-check}
vpc_summary <- sim_vpc |>
  left_join(observed_range, by = "analyte") |>
  group_by(analyte) |>
  summarise(
    median_sim = median(sim),
    q10 = quantile(sim, 0.10), q90 = quantile(sim, 0.90),
    frac_in_observed_range = mean(sim >= lo & sim <= hi),
    .groups = "drop"
  )
knitr::kable(vpc_summary, digits = 2)

# Observed medians read from Figure 1 are about 200 ug/L (riociguat) and
# 110 ug/L (M1). A factor-of-two band leaves room for the 200-subject sampling
# noise (the median's MC-SE is ~5%) and still fails on a unit or
# volume transcription error, which moves concentrations by 10x or more.
stopifnot(
  all(vpc_summary$frac_in_observed_range > 0.75),
  with(vpc_summary, median_sim[analyte == "riociguat"] > 100 &
    median_sim[analyte == "riociguat"] < 400),
  with(vpc_summary, median_sim[analyte == "M1"] > 55 &
    median_sim[analyte == "M1"] < 220)
)
```

# Assumptions and deviations

* **Inter-individual variability.** With one sample per patient the authors
  could not separate inter-individual from residual variability (Methods). The
  model therefore has no etas and the Table 2 proportional variances (0.152 and
  0.268) cover both. They are encoded as SDs (`sqrt(variance)`). Individual
  predictions (`Cc`, `Cc_m1`) are typical values for the given covariates. Use
  the `sim` column for the predictive distribution.
* **Lag-time subgroup.** The 2.95 h lag applies only to the six patients whose
  concentrations were unexpectedly high late after the dose. It is encoded with
  the binary covariate `MIX_LAGGED_ABS`. The paper suggests food as a possible
  cause but cannot confirm it. Set it to 0 for a typical patient.
* **Covariate units.** Total bilirubin enters as the canonical `TBILI` column in
  umol/L. The model converts it to the paper's mg/dL with a factor of 17.1.
  Creatinine clearance (`CRCL`) is used in mL/min as reported. The paper calls
  it a CKD-EPI estimate but does not say whether it was de-normalised from
  mL/min/1.73 m^2.
* **Bilirubin exponent.** Table 2 gives -0.462 and the Results text -0.463. The
  Table 2 final-model value is used.
* **Metabolite mass balance.** The paper does not mention a molecular-weight
  correction for the conversion of riociguat (MW 422.4) to M1 (MW 408.4), so M1
  is formed 1:1 on a mass basis. Correcting for it would lower M1
  concentrations by about 3%. The M1 parameters depend on the assumption that
  the two volumes are equal (Discussion), so absolute M1 parameter values
  should be read with that in mind.
* **ka.** Fixed at 3 1/h from Saleh 2016 (single-dose riociguat popPK). The
  paper labels it 'Ka/F', but a rate constant is not scaled by bioavailability,
  so it is treated as ka.
* **Virtual-cohort covariates.** Log-normal distributions matched to the Table 1
  medians and IQRs, independent of each other. The dose mix is an assumption
  consistent with the Table 1 median and IQR.
* **Errata.** No erratum or correction was found for this article (literature
  check on 2026-09-25).
