---
title: "Mycophenolic acid (Mizaki 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mycophenolic acid (Mizaki 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
options(cli.progress_show_after = Inf)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Mizaki_2023_mycophenolic_acid"))$reference`
* Description: `r rxode2::rxode(readModelDb("Mizaki_2023_mycophenolic_acid"))$description`
* Article: <https://doi.org/10.1186/s40780-022-00271-w> (open access)

Mizaki 2023 developed the first population PK model of mycophenolic acid
(MPA) for Japanese patients with lupus nephritis, from a single 12-h
steady-state profile per patient, and used it for maximum a posteriori
Bayesian limited-sampling strategies. The final model, fitted in Phoenix NLME,
is two-compartment with a six-compartment transit absorption chain and
first-order elimination; Cockcroft-Gault creatinine clearance (CLcr) and serum
albumin are covariates on CL/F. An enterohepatic-circulation model and a lag
time were tested and did not improve the fit.

# Population

Thirty-four adult Japanese patients (6 male, 28 female) receiving mycophenolate
mofetil (MMF, CellCept) 250-1000 mg every 12 h for lupus nephritis at Aichi
Medical University Hospital (March 2015 - June 2022) were studied
retrospectively. Table 1 (median, IQR): age 39.0 years (26.3-51.8), body weight
54.7 kg (47.2-58.6), serum albumin 3.6 g/dL (2.8-3.9), Cockcroft-Gault CLcr
103.6 mL/min (71.1-125.0), MMF 1500 mg/day (1500-1688). All patients received
prednisolone; 8 received tacrolimus, 22 a proton-pump inhibitor and 5
iron/magnesium oxide. Samples were taken at steady state before the dose and
0.5, 1, 2, 3, 4, 6, 8 and 12 h after it (302 concentrations; 33 full profiles),
and measured by enzyme immunoassay (LLOQ 0.40 ug/mL).

The same information is available programmatically via
`rxode2::rxode(readModelDb("Mizaki_2023_mycophenolic_acid"))$population`.

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Two-compartment, transit absorption, first-order elimination, no EHC, no lag | -- | Results 'PPK model development and validation'; Fig. 3 |
| Six transit compartments, `ktr = (n + 1) / MTT` | n = 6 (fixed) | Fig. 3 caption; Results; Discussion 'fixed 6 transit compartments' |
| `lvc` (V1/F) | 22.95 L | Table 2 |
| `lcl` (CL/F at the covariate medians) | 13.15 L/h | Table 2 |
| `lka` (ka) | 2.98 1/h | Table 2 |
| `lvp` (V2/F) | 336.03 L | Table 2 |
| `lq` (Q/F) | 26.44 L/h | Table 2 |
| `lmtt` (MTT) | 0.45 h | Table 2 |
| `e_crcl_cl` | 0.78 | Table 2 'Effect CLcr on CL' |
| `e_alb_cl` | -0.88 | Table 2 'Effect Alb on CL' |
| Power covariate form with medians CLcr 6.2 L/h, Alb 3.6 g/dL | -- | Additional file 2 footnote |
| IIV V1, CL, MTT | 86.96%, 32.23%, 58.50% CV | Table 2; omega^2 = log(1 + CV^2) |
| `propSd` | 0.3409 | Table 2 'Proportional error (CV%)' 34.09 |
| Dose units (mg MMF, no molecular-weight conversion) | -- | Methods (V1/F, CL/F reported as V1, CL) |

# Typical-value checks

```{r model}
mod <- rxode2::rxode(readModelDb("Mizaki_2023_mycophenolic_acid"))
stopifnot(identical(
  mod$state,
  c("depot", paste0("transit", 1:7), "central", "peripheral1")
))
th <- setNames(mod$iniDf$est, mod$iniDf$name)

tau <- 12
# Day-15 morning interval: doses at 0, 12, ..., 336 h, sampling 336-348 h.
# The terminal half-life is about 27 h, so 28 prior doses reach steady state.
n_dose <- 29
t_ss <- (n_dose - 1) * tau
samp <- c(0, 0.5, 1, 2, 3, 4, 6, 8, 12)

make_events <- function(ids, dose, obs_times, n_doses = n_dose) {
  one_id <- function(i, amt) {
    dplyr::bind_rows(
      data.frame(
        id = i, time = seq(0, by = tau, length.out = n_doses),
        amt = amt, evid = 1L, cmt = "depot"
      ),
      data.frame(id = i, time = obs_times, amt = 0, evid = 0L, cmt = "central")
    )
  }
  dose <- rep_len(dose, length(ids))
  dplyr::bind_rows(Map(one_id, ids, dose)) |>
    # Observation before dose at a shared time, so the pre-dose sample is a trough.
    dplyr::arrange(id, time, dplyr::desc(evid == 0L))
}

typ <- function(dose, obs_times, CRCL = 6.2 * 1000 / 60, ALB = 36, n_doses = n_dose) {
  ev <- make_events(1L, dose, obs_times, n_doses)
  ev$CRCL <- CRCL
  ev$ALB <- ALB
  rxode2::rxSolve(mod, ev,
    omega = NA, sigma = NA, returnType = "data.frame",
    rtol = 1e-10, atol = 1e-12, maxsteps = 1e6
  )
}
```

## Covariate effects on CL/F

```{r cl-covariates}
cl_fun <- function(crcl_mlmin, alb_gdl) {
  exp(th[["lcl"]]) * (crcl_mlmin * 0.06 / 6.2)^th[["e_crcl_cl"]] *
    (alb_gdl / 3.6)^th[["e_alb_cl"]]
}
stopifnot(abs(cl_fun(6.2 / 0.06, 3.6) / 13.15 - 1) < 1e-12)

cl_grid <- tidyr::expand_grid(
  crcl = seq(40, 180, by = 2),
  alb = c(2.8, 3.6, 3.9)
) |>
  dplyr::mutate(cl = cl_fun(crcl, alb), alb = factor(paste(alb, "g/dL")))
ggplot(cl_grid, aes(crcl, cl, colour = alb)) +
  geom_line() +
  labs(
    x = "Cockcroft-Gault CLcr (mL/min)", y = "Typical CL/F (L/h)",
    colour = "Serum albumin",
    caption = "Typical CL/F across the Table 1 IQR of albumin (2.8, 3.6, 3.9 g/dL)."
  ) +
  theme_bw()
```

Over the Table 1 interquartile ranges, CL/F runs from about 9 L/h (CLcr 71
mL/min, albumin 3.9 g/dL) to about 19 L/h (CLcr 125 mL/min, albumin 2.8
g/dL): lower renal function raises exposure and hypoalbuminaemia lowers it,
the directions the paper's dose-normalised AUC comparisons (Fig. 2A, B) show.

## Steady-state AUC equals dose / CL

For a linear model the AUC over one steady-state interval equals dose / (CL/F)
regardless of the absorption and distribution parameters.

```{r ss-auc}
trap <- function(x, y) sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)
ss_check <- dplyr::bind_rows(lapply(
  list(c(103.33, 3.6), c(71.1, 3.9), c(125, 2.8)),
  function(cv) {
    s <- typ(750, t_ss + seq(0, tau, by = 0.005), CRCL = cv[1], ALB = cv[2] * 10, n_doses = 61)
    data.frame(
      crcl = cv[1], alb_gdl = cv[2], auc_tau = trap(s$time, s$Cc),
      dose_over_cl = 750 / cl_fun(cv[1], cv[2])
    )
  }
)) |>
  dplyr::mutate(rel_diff = auc_tau / dose_over_cl - 1)
knitr::kable(ss_check, digits = 4)
# 61 doses (30 days) leave < 1e-4 of steady state unreached, and trapezoid
# error on a 0.005-h grid is ~1e-5; a wrong CL, covariate form or dose unit
# moves this by tens of percent.
stopifnot(all(abs(ss_check$rel_diff) < 2e-3))
```

## Typical steady-state profile

```{r typical-profile}
prof <- typ(750, t_ss + seq(0, 12, by = 0.02))
tmax_typ <- prof$time[which.max(prof$Cc)] - t_ss
ggplot(prof, aes(time - t_ss, Cc)) +
  geom_line() +
  labs(
    x = "Time after the morning dose (h)", y = "MPA (mg/L)",
    caption = "Typical patient (CLcr 103 mL/min, albumin 3.6 g/dL), MMF 750 mg every 12 h at steady state."
  ) +
  theme_bw()
tmax_typ
# Mean arrival into the absorption compartment is MTT = 0.45 h, then ka = 2.98
# 1/h; the observed median Tmax is 1.0 h (IQR 1.0-1.0).
stopifnot(tmax_typ > 0.5, tmax_typ < 1.5)
```

The typical profile has a single peak. In 32 of the 33 full profiles the
patients showed a second peak at a median 8 h (median 3.3 ug/mL), which the
paper attributes to enterohepatic circulation; the published final model has
no EHC term, so neither does this one.

# Virtual cohort and NCA against the observed medians

Two hundred virtual patients. CLcr and albumin are drawn from piecewise-uniform
distributions that reproduce the Table 1 quartiles exactly (the minimum and
maximum, which the paper does not report, are set by the maintainers to 40 and
180 mL/min and 1.8 and 4.6 g/dL). The MMF dose is 500, 750 or 1000 mg every
12 h with probabilities 0.10, 0.65 and 0.25, which reproduces the Table 1
median of 1500 mg/day with an upper quartile between 1500 and 2000 mg/day. The
paper's nine-point schedule is simulated with residual error, and AUC0-12 uses
the linear trapezoidal rule as in the paper's NCA.

```{r cohort}
set.seed(20230109)
n_sub <- 200
r_quartiles <- function(n, q) {
  # q = c(min, Q1, median, Q3, max); each quartile band gets 25% of draws.
  band <- sample.int(4, n, replace = TRUE)
  stats::runif(n, q[band], q[band + 1])
}
cohort <- data.frame(
  id = seq_len(n_sub),
  CRCL = r_quartiles(n_sub, c(40, 71.1, 103.6, 125.0, 180)),
  ALB = 10 * r_quartiles(n_sub, c(1.8, 2.8, 3.6, 3.9, 4.6)),
  dose = sample(c(500, 750, 1000), n_sub, replace = TRUE, prob = c(0.10, 0.65, 0.25))
)
ev_cohort <- make_events(cohort$id, cohort$dose, t_ss + samp) |>
  dplyr::left_join(dplyr::select(cohort, id, CRCL, ALB), by = "id")

rxode2::rxSetSeed(20230109)
sim <- rxode2::rxSolve(mod, ev_cohort, returnType = "data.frame", maxsteps = 1e6) |>
  dplyr::select(-dplyr::any_of(c("CRCL", "ALB"))) |>
  dplyr::left_join(cohort, by = "id") |>
  dplyr::mutate(treatment = "MMF bid, steady state", tad = time - t_ss)
stopifnot(!anyNA(sim$sim), nrow(sim) == n_sub * length(samp))
```

```{r vpc, fig.height = 4}
sim |>
  dplyr::group_by(tad) |>
  dplyr::summarise(
    p05 = quantile(sim, 0.05), p50 = median(sim), p95 = quantile(sim, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.25) +
  geom_line() +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)", y = "MPA (mg/L)",
    caption = "Simulated 5th-50th-95th percentiles with residual error; compare with Figure 1F of Mizaki 2023."
  ) +
  theme_bw()
```

```{r nca}
conc_df <- sim |>
  dplyr::transmute(id, treatment, time = tad, conc = sim) |>
  dplyr::filter(!is.na(conc))
# A normal proportional error makes Cc * (1 + eps) negative when eps < -1
# (P = 0.0017 for sd 0.34). Such samples would read as below the assay limit,
# so they enter the NCA as zero.
stopifnot(mean(conc_df$conc < 0) < 0.02)
conc_df$conc <- pmax(conc_df$conc, 0)
dose_df <- cohort |>
  dplyr::mutate(treatment = "MMF bid, steady state", time = 0) |>
  dplyr::select(id, treatment, time, dose)

intervals <- data.frame(
  start = 0, end = 12, cmax = TRUE, tmax = TRUE, auclast = TRUE
)
o_conc <- PKNCA::PKNCAconc(conc_df, conc ~ time | treatment + id)
o_dose <- PKNCA::PKNCAdose(dose_df, dose ~ time | treatment + id)
o_data <- PKNCA::PKNCAdata(o_conc, o_dose,
  intervals = intervals,
  options = list(auc.method = "linear")
)
nca_res <- PKNCA::pk.nca(o_data)
nca_df <- as.data.frame(nca_res)

# The paper reports medians (Results 'Pharmacokinetic analysis'); C0 is the
# pre-dose sample, which at steady state is the trough PKNCA reads at time 0.
c0 <- conc_df |>
  dplyr::filter(time == 0) |>
  dplyr::summarise(PPORRES = median(conc)) |>
  dplyr::mutate(treatment = "MMF bid, steady state", PPTESTCD = "c0")
sim_median <- nca_df |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast")) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(PPORRES = median(PPORRES), .groups = "drop")

published <- data.frame(
  treatment = "MMF bid, steady state",
  cmax = 16.7, tmax = 1.0, auclast = 51.1
)
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_median,
  reference = published,
  by = "treatment",
  units = c(cmax = "mg/L", tmax = "h", auclast = "mg*h/L"),
  tolerance_pct = 20
)
knitr::kable(cmp,
  caption = paste(
    "Simulated (median of 200 virtual patients) vs the observed medians of",
    "Mizaki 2023 (Results). * differs from the reference by more than 20%."
  )
)
attr(cmp, "footnote")

data.frame(
  quantity = "C0 (pre-dose) median, mg/L",
  simulated = round(c0$PPORRES, 2),
  published = 1.8
) |>
  dplyr::rename("Quantity" = quantity, "Simulated" = simulated, "Published" = published) |>
  knitr::kable()
```

```{r nca-gate}
get_sim <- function(p) {
  v <- sim_median$PPORRES[sim_median$PPTESTCD == p]
  if (length(v) != 1L) stop("no unique simulated value for ", p)
  v
}
auc_pct <- 100 * (get_sim("auclast") / 51.1 - 1)
cmax_pct <- 100 * (get_sim("cmax") / 16.7 - 1)
auc_pct
cmax_pct
stopifnot(
  # AUC0-12 is dose / CL-dominated: a wrong clearance, covariate centre or
  # dose unit moves it by tens of percent. The cohort's dose and covariate
  # mix is approximate (+12% on one 2-thread run), and the median of 200
  # subjects has a Monte Carlo SE of about 3.5%, so 25 leaves ~3.5 SE of
  # headroom.
  abs(auc_pct) < 25,
  # Cmax depends on MTT, ka and V1 (CV 87%), so the band is wider.
  abs(cmax_pct) < 35,
  # Observed median Tmax 1.0 h on the 0.5 / 1 / 2 h grid; a median over an
  # even count can fall between two grid points.
  get_sim("tmax") >= 0.5, get_sim("tmax") <= 2
)
```

The simulated median Cmax and Tmax match the observed medians, and AUC0-12 is
about 12% above the observed 51.1 mg h/L, within what the approximate dose and
covariate mix of the virtual cohort allows (the typical patient on 750 mg every
12 h has AUC0-12 = 750 / 13.15 = 57.0 mg h/L).

The pre-dose (trough) concentration is the one quantity the model does not
reproduce: the simulated median is roughly 50% above the observed 1.8 mg/L.
This is a property of the published model rather than of the transcription,
and the paper itself shows it. In Table 3 the Bayesian strategy that uses only
the C0 sample is the one single-sample strategy whose mean prediction error
excludes zero (-11.86%, 95% CI -18.23 to -5.48): a model that predicts a
higher trough for a given AUC than patients actually have will underestimate
AUC from an observed trough. The final model has no enterohepatic-circulation
term, so the late-interval concentrations are carried by a large, slowly
equilibrating peripheral compartment (V2/F 336 L, terminal half-life about 27
h), which is one plausible reason for the elevated trough. The trough is
therefore reported here but not gated.

# Assumptions and deviations

- **Covariate form and centring.** The main text reports only the exponents
  (Table 2). The power form and the centring values, CLcr 6.2 L/h and albumin
  3.6 g/dL, are printed under the parameter table of the four-covariate
  candidate model in Additional file 2. The final model is that model without
  the PPI and iron/magnesium oxide terms on V1, fitted to the same data, so the
  same form and medians are applied here. Table 1 gives the median CLcr as
  103.6 mL/min (6.22 L/h); the model uses the printed 6.2 L/h (103.3 mL/min).
- **CLcr units.** The model was fitted with CLcr in L/h. The canonical `CRCL`
  column is the raw (not BSA-normalised) Cockcroft-Gault value in mL/min and is
  converted to L/h inside the model.
- **Albumin units.** The canonical `ALB` column is in g/L; the model converts to
  g/dL so the published exponent applies unchanged.
- **Transit chain.** Fig. 3 draws six boxes (a1-a6) with the dose entering a1,
  while its caption defines `ktr = (n + 1) / MTT` with n = 6 transit
  compartments. The maintainers followed the equation: the Savic form that
  Phoenix NLME's transit statement implements has a dose compartment plus n
  transit compartments draining at ktr, so the mean arrival time into the
  absorption compartment equals MTT. The model encodes this as `depot` (dose)
  plus `transit1`-`transit6`, with `transit7` as the absorption compartment
  emptying into `central` at ka. Reading the drawing literally (six ktr steps)
  would shorten the mean absorption delay from 0.45 to 0.39 h.
- **IIV.** Table 2 reports IIV as CV% without saying how it was computed from
  the Phoenix omega estimates; the variances here are omega^2 =
  log(1 + CV^2). Nothing in the paper separates this from the alternative
  reading CV% = 100 x omega, which matters mainly for V1/F (0.563 here versus
  0.756) and little for CL/F (0.099 versus 0.104). No covariances are reported.
- **No enterohepatic circulation.** The published final model has none, so the
  second peak observed in 32 of 33 profiles is not reproduced, and the
  simulated trough runs about 50% above the observed median (see the NCA
  section; the paper's own Table 3 C0-only Bayesian bias points the same way).
- **Virtual cohort.** Covariate extremes and the dose mix are maintainers'
  choices made to match the Table 1 quartiles; the paper does not report them.
- **Errata.** No correction notice was found in Europe PMC or Crossref as of
  2026-10-10.
