---
title: "Shikimic acid PK and immune-enhancing PD in growing pigs (Mo 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shikimic acid PK and immune-enhancing PD in growing pigs (Mo 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Shikimic acid (SA) is a trihydroxycyclohexene carboxylic acid used as a food
preservative and as the synthetic precursor of oseltamivir. Mo 2024 is the first
report of its pharmacokinetics in pigs. Six growing pigs received SA in a
two-period, two-sequence crossover: 50 mg/kg by gavage in one period and
2 mg/kg into the marginal ear vein in the other, separated by a 7-day washout.
Plasma SA was measured by HPLC-MS/MS, and complement C3 and C4 and
immunoglobulin G, A and M were measured by ELISA in the same samples.

The authors selected **different structural models for the two routes** by AIC
(Section 2.6, supplementary Table S2) and fitted each pig individually in
Phoenix WinNonlin, so the paper contributes **two** independent model files:

```{r models}
sa_models <- c(
  "Mo_2024_shikimicAcid_pig_iv",    # 2-compartment IV disposition + sigmoid-Emax PD
  "Mo_2024_shikimicAcid_pig_oral"   # 1-compartment first-order absorption
)

# Parse each model once; the parsed object is what exposes the file-level
# metadata (reference, description, population) as well as being solvable.
mod_iv   <- rxode2::rxode(readModelDb(sa_models[1]))
mod_oral <- rxode2::rxode(readModelDb(sa_models[2]))
```

The immune-response (PD) layer belongs to the intravenous model only: Mo 2024
fitted sigmoid-Emax models to the intravenous group because the intragastric
effect-versus-concentration curves were not monotonic (Section 4, Figures S4 and
S5).

* Citation: `r mod_iv$reference`
* Article: [J Agric Food Chem 2024;72:26224-26235](https://doi.org/10.1021/acs.jafc.4c09250)

```{r description}
sa_parsed <- stats::setNames(list(mod_iv, mod_oral), sa_models)
for (m in sa_models) {
  cat(m, "\n", sep = "")
  cat(strwrap(sa_parsed[[m]]$description, 76), sep = "\n")
  cat("\n\n")
}
```

# Population

Six growing pigs (Landrace x Large White, approximately 90 days old, both
sexes) of comparable health status and genetic background were randomised to
sequence A (n = 3) or sequence B (n = 3) after a 7-day acclimatisation
(Mo 2024 Section 2.2, Table 1). Every pig contributed both an intragastric and
an intravenous profile, so the same six animals support both model files.
Plasma was drawn from the anterior vena cava predose and at 0.083 (intravenous
only), 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 12 and 24 h; complement and
immunoglobulin were assayed at 0, 0.5, 1, 1.5, 3, 6 and 12 h (Sections 2.3 and
2.5).

Body weights are reported only in supplementary Table S1, which is not part of
the open-access record; the paper states that weight did not differ between the
two periods (P = 0.320). Because every parameter in both models is expressed per
kg body weight, no weight covariate is needed to simulate them.

```{r population}
str(mod_iv$population)
```

# Source trace

Both model files carry the origin of each value as an in-file comment next to
its `ini()` entry. The table below collects them.

| Equation / parameter | Value | Source location |
|---|---|---|
| **`Mo_2024_shikimicAcid_pig_iv`** | | |
| IV disposition equation `C = A*exp(-alpha*t) + B*exp(-beta*t)` | n/a | Section 3.2, second paragraph |
| `lvc` (central volume) | 244.81 mL/kg | derived: `Dose/(A + B)` from Table 10 means |
| `lcl` (clearance) | 229.14 mL/h/kg | derived: `k10*vc` from Table 10 means |
| `lq` (intercompartmental clearance) | 115.23 mL/h/kg | derived: `k12*vc` from Table 10 means |
| `lvp` (peripheral volume) | 300.58 mL/kg | derived: `q/k21` from Table 10 means |
| PD equation `E = Emax*C^gamma/(EC50^gamma + C^gamma)` | n/a | eq 2, Section 2.6 |
| `lemax_c3` / `lec50_c3` / `lhill_c3` | 352.95 ug/mL / 700.45 ng/mL / 3.77 | Table 11, row C3 |
| `lemax_c4` / `lec50_c4` / `lhill_c4` | 0.064 ug/mL / 836.61 ng/mL / 3.89 | Table 11, row C4 |
| `lemax_igg` / `lec50_igg` / `lhill_igg` | 9801.86 ug/mL / 514.22 ng/mL / 5.15 | Table 11, row IgG |
| `lemax_iga` / `lec50_iga` / `lhill_iga` | 469.30 ug/mL / 561.88 ng/mL / 6.97 | Table 11, row IgA |
| `lemax_igm` / `lec50_igm` / `lhill_igm` | 1252.05 ug/mL / 686.24 ng/mL / 3.75 | Table 11, row IgM |
| `addSd`, `addSd_dC3` ... `addSd_dIgM` | 0 (fixed) | not reported; see Errata |
| **`Mo_2024_shikimicAcid_pig_oral`** | | |
| Absorption equation `C = M*(exp(-ke*t) - exp(-ka*t))` | n/a | Section 3.2, first paragraph (sign corrected; see Errata) |
| `lka` | 0.815467 /h | derived: `ln(2)/0.85 h` from Table 9 `t1/2ka` |
| `lcl` (apparent Cl/F) | 1086.52 mL/h/kg | Table 9, `Cl_F` |
| `lvc` (apparent V/F) | 2837.21 mL/kg | derived: `Cl_F/(ln(2)/1.81 h)` from Table 9 |
| `addSd` | 0 (fixed) | not reported; see Errata |

## Reconstructing the disposition parameters from Table 10

Mo 2024 tabulates the intravenous fit as the biexponential coefficients and
exponents rather than as volumes and clearances, so the packaged parameters are
a transformation of Table 10. The chunk below performs that transformation and
solves it back, confirming that the packaged values regenerate Table 10.

```{r derive-iv}
A <- 7227.76; B <- 941.91; alpha <- 1.56; beta <- 0.23   # Table 10, mean column
dose_iv <- 2e6                                           # 2 mg/kg expressed as ng/kg

vc  <- dose_iv / (A + B)
k21 <- (A * beta + B * alpha) / (A + B)
k10 <- alpha * beta / k21
k12 <- alpha + beta - k21 - k10
cl  <- k10 * vc
q   <- k12 * vc
vp  <- q / k21

# Round as packaged, then solve the rounded values back to a biexponential.
vcr <- round(vc, 2); clr <- round(cl, 2); qr <- round(q, 2); vpr <- round(vp, 2)
K10 <- clr / vcr; K12 <- qr / vcr; K21 <- qr / vpr
s <- K10 + K12 + K21; p <- K10 * K21
Al <- (s + sqrt(s^2 - 4 * p)) / 2
Be <- (s - sqrt(s^2 - 4 * p)) / 2

data.frame(
  quantity  = c("A (ng/mL)", "alpha (1/h)", "B (ng/mL)", "beta (1/h)"),
  published = c(A, alpha, B, beta),
  recovered = round(c(dose_iv / vcr * (Al - K21) / (Al - Be), Al,
                      dose_iv / vcr * (K21 - Be) / (Al - Be), Be), 4)
) |>
  dplyr::rename("Table 10 quantity" = quantity,
                "Published" = published,
                "Recovered from packaged model" = recovered) |>
  knitr::kable(caption = "The packaged vc/cl/q/vp regenerate Table 10 to rounding.")
```

Mo 2024 also tabulates secondary parameters computed independently of the
biexponential, which act as an external check on the transformation:

```{r derive-crosscheck}
data.frame(
  parameter = c("Cl (mL/h/kg)", "Vss (mL/kg)", "peripheral volume (mL/kg)"),
  published = c(233.20, 574.10, 323.11),
  derived   = round(c(cl, vc + vp, vp), 2)
) |>
  dplyr::mutate("Difference (%)" = round(100 * (derived - published) / published, 1)) |>
  dplyr::rename("Parameter" = parameter,
                "Mo 2024 Table 10" = published,
                "Derived here" = derived) |>
  knitr::kable(caption = paste(
    "Secondary-parameter cross-check. Differences arise because Table 10 reports",
    "the mean of six individually derived values whereas the packaged model",
    "transforms the mean biexponential; the two are not algebraically identical."
  ))
```

### The `V_Z` column of Table 10 is the peripheral volume

Table 10 labels a column `V_Z` (mean 323.11 mL/kg) and the abbreviation list
glosses it as "the apparent volume of distribution". Taken as the terminal-phase
volume it is irreconcilable with the rest of the table: `Cl/beta` runs from 660
to 1913 mL/kg across the six pigs (mean 1115), i.e. 2.3 to 4.9 times the
tabulated `V_Z` of the same animal. The column is in fact the
**peripheral** volume, because `Vss - V_Z` equals `Dose/(A + B)` -- the central
volume -- for every one of the six pigs:

```{r vz-identity}
pigs <- data.frame(
  A    = c(5520.89, 6305.87, 7713.37, 6992.87, 7849.21, 8984.33),
  B    = c(1115.22,  357.49, 1181.81,  590.59, 1658.52,  747.84),
  Vss  = c( 794.73,  777.21,  486.09,  581.23,  417.74,  387.64),
  Vz   = c( 493.34,  477.06,  261.25,  317.50,  207.38,  182.13),
  Cl   = c( 125.28,  382.68,  271.04,  174.31,  250.80,  195.07),
  beta = c(   0.11,    0.20,    0.32,    0.14,    0.38,    0.22),
  thalf_beta = c(6.41, 3.42, 2.16, 5.01, 1.84, 3.11)
)
pigs |>
  dplyr::transmute(
    pig            = seq_len(dplyr::n()),
    `Dose/(A+B)`   = round(2e6 / (A + B), 2),
    `Vss - V_Z`    = round(Vss - Vz, 2),
    `Cl/beta`      = round(Cl / beta, 0)
  ) |>
  knitr::kable(caption = paste(
    "Vss - V_Z reproduces the central volume Dose/(A+B) to within 0.01 mL/kg for",
    "all six pigs, whereas the terminal-phase volume Cl/beta does not match V_Z",
    "at all."
  ))

stopifnot(max(abs(2e6 / (pigs$A + pigs$B) - (pigs$Vss - pigs$Vz))) < 0.02)
```

The packaged `lvp` therefore corresponds to that column, and no terminal-phase
volume is carried.

# Virtual cohort

Mo 2024 fitted each animal separately and reported only the mean and SD of the
individual estimates, so neither model carries between-subject variability or a
residual-error model (see Errata). Both simulations below are therefore
deterministic single typical-animal profiles at the studied dose levels, on the
paper's own sampling schedule augmented with a dense grid for plotting.

```{r cohort}
obs_iv   <- c(0, 0.083, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 12)
obs_oral <- c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 12, 24)
obs_pd   <- c(0, 0.5, 1, 1.5, 3, 6, 12)

grid_iv   <- sort(unique(c(obs_iv,   obs_pd, seq(0, 12, length.out = 241))))
# The intragastric grid is refined around the absorption peak so that the NCA
# T_max reflects the model rather than the spacing of the observation grid.
grid_oral <- sort(unique(c(obs_oral, seq(0, 24, length.out = 241),
                          seq(1, 2.5, by = 0.01))))

# Intravenous: six algebraic endpoints (Cc plus five immune markers), so
# observation rows carry dvid = 1 and no cmt. rxode2 returns every algebraic
# observable as a column regardless of which endpoint the row names.
ev_iv <- as.data.frame(
  rxode2::et(amt = dose_iv, cmt = "central") |>
    rxode2::et(grid_iv)
)
ev_iv$dvid <- ifelse(ev_iv$evid == 0, 1L, NA_integer_)
ev_iv$id <- 1L

# Intragastric: single endpoint, so observations sit on the central ODE state.
dose_oral <- 5e7   # 50 mg/kg expressed as ng/kg
ev_oral <- as.data.frame(
  rxode2::et(amt = dose_oral, cmt = "depot") |>
    rxode2::et(grid_oral, cmt = "central")
)
ev_oral$id <- 1L

c(iv_rows = nrow(ev_iv), oral_rows = nrow(ev_oral))
```

# Simulation

```{r simulate}
sim_iv <- rxode2::rxSolve(
  mod_iv, ev_iv,
  returnType = "data.frame", addDosing = FALSE
)
sim_oral <- rxode2::rxSolve(
  mod_oral, ev_oral,
  returnType = "data.frame", addDosing = FALSE
)

stopifnot(nrow(sim_iv) > 0, nrow(sim_oral) > 0)
stopifnot(all(c("Cc", "dC3", "dC4", "dIgG", "dIgA", "dIgM") %in% names(sim_iv)))
stopifnot(!anyNA(sim_iv$Cc), !anyNA(sim_oral$Cc))
```

# Replicate published figures

## Figure 3B -- intravenous concentration-time profile

The simulated curve is overlaid on the paper's own analytic solution
`C = A*exp(-alpha*t) + B*exp(-beta*t)` evaluated at the Table 10 means, and on
the observed mean C_max.

```{r fig3b, fig.cap = "Replicates Figure 3B of Mo 2024 (intravenous, 2 mg/kg)."}
# Evaluated at the solver's own output times so the overlay and the numeric
# check below are aligned by construction.
analytic_iv <- data.frame(time = sim_iv$time) |>
  dplyr::mutate(Cc = A * exp(-alpha * time) + B * exp(-beta * time))

ggplot(sim_iv, aes(time, Cc)) +
  geom_line(linewidth = 1) +
  geom_line(data = analytic_iv, aes(time, Cc),
            colour = "firebrick", linetype = "22", linewidth = 0.9) +
  geom_point(data = data.frame(time = 0, Cc = 8169.67),
             aes(time, Cc), colour = "firebrick", size = 2.5) +
  geom_hline(yintercept = 20, linetype = "dotted", colour = "grey40") +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Shikimic acid (ng/mL, log scale)",
    subtitle = paste("Black: packaged model. Red dashed: Mo 2024 analytic biexponential.",
                     "\nRed point: reported mean C_max. Dotted line: LOQ 20 ng/mL.")
  ) +
  theme_bw()
```

```{r fig3b-check}
max_rel_err <- max(abs(sim_iv$Cc - analytic_iv$Cc) / analytic_iv$Cc)
cat(sprintf("Largest relative deviation from the published equation: %.4f%%\n",
            100 * max_rel_err))
stopifnot(max_rel_err < 0.01)
```

## Figure 3A -- intragastric concentration-time profile

```{r fig3a, fig.cap = "Replicates Figure 3A of Mo 2024 (intragastric, 50 mg/kg)."}
ggplot(sim_oral, aes(time, Cc)) +
  geom_line(linewidth = 1) +
  geom_point(data = data.frame(time = 1.78, Cc = 10823.44),
             aes(time, Cc), colour = "firebrick", size = 2.5) +
  geom_errorbar(
    data = data.frame(time = 1.78, lo = 10823.44 - 4934.04, hi = 10823.44 + 4934.04),
    aes(x = time, ymin = lo, ymax = hi), inherit.aes = FALSE,
    width = 0.4, colour = "firebrick"
  ) +
  labs(
    x = "Time (h)", y = "Shikimic acid (ng/mL)",
    subtitle = paste("Black: packaged model. Red: reported mean C_max at mean T_max",
                     "(+/- SD of C_max), Table 9.")
  ) +
  theme_bw()
```

## Figure 5 -- immune-marker time courses after intravenous dosing

The PD layer predicts the **absolute change from the predose level**, which is
what Mo 2024's eq 2 models. Figure 5 plots absolute concentrations; the paper's
baselines are not tabulated, so the change is plotted directly.

One difference from Figure 5 is structural rather than a discrepancy. In the
model, `t = 0` is the instant *after* the bolus, where the concentration is
already at C_max; because eq 2 is a direct-effect model with no effect
compartment, the predicted change is therefore already near E_max at `t = 0`.
In the paper, the 0 h ELISA sample is the *predose* baseline, against which the
change is defined as zero, and the first post-dose measurement is at 0.5 h. The
model consequently cannot show the 0-to-0.5 h rise that Figure 5 displays: a
direct-effect model driven by a bolus has no mechanism to delay the onset. Mo
2024's own finding of no hysteresis is what licenses that structure. The two
testable features among the **post-dose** samples are that the effect is maximal
at the earliest one (0.5 h) and has returned to the predose level by 12 h.

```{r fig5, fig.cap = "Replicates the time-course shape of Figure 5 of Mo 2024 (intravenous)."}
pd_long <- sim_iv |>
  dplyr::select(time, dC3, dC4, dIgG, dIgA, dIgM) |>
  tidyr::pivot_longer(-time, names_to = "marker", values_to = "delta") |>
  dplyr::mutate(marker = factor(
    marker,
    levels = c("dC3", "dC4", "dIgG", "dIgA", "dIgM"),
    labels = c("C3", "C4", "IgG", "IgA", "IgM")
  ))

ggplot(pd_long, aes(time, delta)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = c(0.5, 12), linetype = "dotted", colour = "grey40") +
  facet_wrap(~marker, scales = "free_y") +
  labs(x = "Time (h)", y = "Change from predose (ug/mL)",
       subtitle = "Dotted lines: the 0.5 h peak and the 12 h return to predose reported by Mo 2024.") +
  theme_bw()
```

```{r fig5-check}
pd_at <- function(t) {
  sim_iv[which.min(abs(sim_iv$time - t)), c("dC3", "dC4", "dIgG", "dIgA", "dIgM")]
}
emax_pub <- c(dC3 = 352.95, dC4 = 0.064, dIgG = 9801.86, dIgA = 469.30, dIgM = 1252.05)

data.frame(
  marker      = names(emax_pub),
  emax        = unname(emax_pub),
  at_0.5h     = as.numeric(pd_at(0.5)),
  at_12h      = as.numeric(pd_at(12))
) |>
  dplyr::mutate(
    `Percent of Emax at 0.5 h` = round(100 * at_0.5h / emax, 1),
    `Percent of Emax at 12 h`  = round(100 * at_12h  / emax, 2)
  ) |>
  dplyr::select(-at_0.5h, -at_12h) |>
  dplyr::rename("Marker" = marker, "Published Emax (ug/mL)" = emax) |>
  knitr::kable(caption = paste(
    "The predicted effect is essentially maximal at the 0.5 h ELISA sample and",
    "has decayed to under 0.1% of Emax by 12 h, matching Mo 2024 Figure 5 and",
    "the statement that levels return to the 0 h value at 12 h."
  ))

# Among the POST-DOSE ELISA samples the effect must be maximal at the earliest
# one (0.5 h) and negligible by 12 h, for every marker. t = 0 is excluded: in the
# model it is the post-bolus instant (change already near Emax), whereas in the
# paper it is the predose baseline where the change is zero by definition. Rows
# are matched to the nominal ELISA times by nearest neighbour rather than by
# floating-point equality.
obs_pd_post <- obs_pd[obs_pd > 0]
pd_obs <- sim_iv[vapply(obs_pd_post, function(t) which.min(abs(sim_iv$time - t)), 1L), ]
for (mk in names(emax_pub)) {
  stopifnot(isTRUE(all.equal(pd_obs$time[which.max(pd_obs[[mk]])], 0.5)))
  stopifnot(pd_at(0.5)[[mk]] / emax_pub[[mk]] > 0.95)
  stopifnot(pd_at(12)[[mk]] / emax_pub[[mk]] < 0.01)
}
```

## Table 11 -- EC50 relative to C_max

Table 11 reports `EC50/C_max` as a percentage for each marker. Because the
packaged model reproduces the intravenous C_max exactly (`A + B` = 8169.67
ng/mL), this column is a direct check on the packaged EC50 values.

```{r ec50-cmax}
ec50_pub <- c(C3 = 700.45, C4 = 836.61, IgG = 514.22, IgA = 561.88, IgM = 686.24)
cmax_model <- max(sim_iv$Cc)

data.frame(
  marker    = names(ec50_pub),
  published = c(8.97, 10.64, 6.56, 7.56, 9.18),
  model     = round(100 * unname(ec50_pub) / cmax_model, 2)
) |>
  dplyr::rename("Marker" = marker,
                "Mo 2024 Table 11 EC50/Cmax (%)" = published,
                "Packaged model EC50/Cmax (%)" = model) |>
  knitr::kable(caption = paste(
    "Mo 2024 reports the mean of six individual ratios; the packaged model forms",
    "the ratio of the mean EC50 to the mean C_max, so small differences are",
    "expected. Both agree that every marker except C4 sits below 10%."
  ))
```

# PKNCA validation

NCA is run on each simulated profile over the window the paper could quantify
(0-12 h intravenous, where SA fell below the 20 ng/mL LOQ beyond 12 h;
0-24 h intragastric).

```{r pknca}
nca_input <- dplyr::bind_rows(
  sim_iv   |> dplyr::select(time, Cc) |> dplyr::mutate(id = 1L, treatment = "IV 2 mg/kg"),
  sim_oral |> dplyr::select(time, Cc) |> dplyr::mutate(id = 2L, treatment = "Intragastric 50 mg/kg")
) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::arrange(treatment, id, time)

stopifnot(all(tapply(nca_input$time, nca_input$id, min) == 0))

conc_obj <- PKNCA::PKNCAconc(nca_input, Cc ~ time | treatment + id)

dose_df <- data.frame(
  id        = c(1L, 2L),
  time      = c(0, 0),
  amt       = c(dose_iv, dose_oral),
  treatment = c("IV 2 mg/kg", "Intragastric 50 mg/kg")
)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE, cl.obs = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

## Comparison against published NCA

```{r nca-compare}
# Long form, so the intravenous bolus (whose T_max is 0 by construction and
# which Mo 2024 does not tabulate) simply contributes no tmax row.
published_nca <- tibble::tribble(
  ~treatment,              ~PPTESTCD,     ~PPORRES,
  "IV 2 mg/kg",            "cmax",         8169.67,   # Table 10
  "IV 2 mg/kg",            "aucinf.obs",   9711.70,   # Table 10
  "IV 2 mg/kg",            "half.life",       3.66,   # Table 10, t1/2beta
  "IV 2 mg/kg",            "cl.obs",        233.20,   # Table 10, Cl
  "Intragastric 50 mg/kg", "cmax",        10823.44,   # Table 9
  "Intragastric 50 mg/kg", "tmax",            1.78,   # Table 9
  "Intragastric 50 mg/kg", "aucinf.obs",  50083.87,   # Table 9
  "Intragastric 50 mg/kg", "half.life",       1.81,   # Table 9, t1/2ke
  "Intragastric 50 mg/kg", "cl.obs",       1086.52    # Table 9, Cl_F
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published_nca,
  by            = "treatment",
  params        = c("cmax", "tmax", "aucinf.obs", "half.life", "cl.obs"),
  units         = c(cmax = "ng/mL", tmax = "h", aucinf.obs = "h*ng/mL",
                    half.life = "h", cl.obs = "mL/h/kg"),
  tolerance_pct = 20
)

knitr::kable(cmp, caption = paste(
  "Simulated versus published NCA. Mo 2024 Table 10 (intravenous) and Table 9",
  "(intragastric). * marks a difference above 20%."
))

# No row may exceed the 20% tolerance; ncaComparisonTable marks those with "*".
stopifnot(!any(grepl("*", cmp[[ncol(cmp)]], fixed = TRUE)))
```

Every parameter agrees within 20% (asserted above, not merely claimed). Two of
the gaps deserve a note:

* **Intravenous `half.life`.** The packaged model carries the mean
  `beta = 0.23 /h`, so its terminal half-life is `ln(2)/0.23` = 3.01 h (PKNCA
  recovers 2.99 h from the simulated profile), whereas Table 10 reports 3.66 h
  -- the *arithmetic mean of the six individual half-lives*. Because the
  half-life is the reciprocal of the exponent, the mean of the half-lives
  exceeds the half-life of the mean exponent. Averaging Table 10's own
  `t1/2beta` row returns exactly the reported 3.66 h; recomputing each
  half-life as `ln(2)/beta_i` from the rounded betas returns 3.64 h; and the
  half-life of the *mean* of those betas is 3.04 h, essentially the 3.01 h the
  packaged model carries. All of these are correct summaries of the same six
  fits, and this -18% gap is the largest in the comparison table.
* **Intragastric `cmax`.** The packaged model gives 9025 ng/mL against a
  reported mean of 10,823 +/- 4934 ng/mL. C_max is not a linear function of the
  parameters, so the C_max of the mean-parameter profile is not the mean of the
  six observed C_max values; the model value sits well inside one SD.

```{r nca-halflife-note}
betas <- pigs$beta
c(`mean of the tabulated t1/2beta (h)`     = round(mean(pigs$thalf_beta), 2),
  `mean of ln(2)/beta_i, rounded betas (h)` = round(mean(log(2) / betas), 2),
  `half-life of the mean beta (h)`          = round(log(2) / mean(betas), 2))
```

## Absolute bioavailability

Mo 2024 reports `F = 21.68%` from the AUC ratio (eq 1). Three independent routes
to `F` -- two from the packaged models, one from the published mean AUCs -- agree
with it.

```{r bioavailability}
nca_df <- as.data.frame(nca_res)
get_nca <- function(trt, param) {
  v <- nca_df$PPORRES[nca_df$treatment == trt & nca_df$PPTESTCD == param]
  stopifnot(length(v) == 1L)
  v
}
auc_iv   <- get_nca("IV 2 mg/kg", "aucinf.obs")
auc_oral <- get_nca("Intragastric 50 mg/kg", "aucinf.obs")

data.frame(
  route = c(
    "Mo 2024 eq 1 (reported)",
    "AUC ratio of the two simulated profiles",
    "Ratio of intravenous Cl to intragastric Cl/F",
    "AUC ratio of the published mean AUCs (Tables 9 and 10)"
  ),
  F_pct = round(c(
    21.68,
    100 * (auc_oral / auc_iv) * (dose_iv / dose_oral),
    100 * 229.14 / 1086.52,
    100 * (50083.87 / 9711.70) * (2 / 50)
  ), 2)
) |>
  dplyr::rename("Estimate of F" = route, "F (%)" = F_pct) |>
  knitr::kable(caption = paste(
    "The packaged models reproduce the reported absolute bioavailability to",
    "about one percentage point by three independent routes."
  ))
```

# Assumptions and deviations

* **The intragastric absorption equation is printed with the wrong sign.**
  Section 3.2 gives `C = M*(exp(-ke*t) + exp(-ka*t))`. A sum of two decaying
  exponentials falls monotonically from `2M` at `t = 0` and has no T_max at all,
  which contradicts the reported T_max of 1.78 h and Figure 3A. The model
  encodes the Bateman difference form `C = M*(exp(-ke*t) - exp(-ka*t))`, whose
  analytic T_max from the packaged `ka` and `ke` is
  `ln(ka/ke)/(ka - ke) = 1.75 h` -- within 2% of the reported mean. Section 3.2
  also introduces the two-compartment model with "following intragastric
  administration" where the data are intravenous; both appear to be
  typographical slips.
* **The `V_Z` column of Table 10 is the peripheral volume, not the
  terminal-phase volume.** Demonstrated in the source-trace section above:
  `Vss - V_Z` equals `Dose/(A + B)` for all six pigs to within 0.01 mL/kg, while
  `Cl/beta` is 2.3 to 4.9 times `V_Z`. `lvp` is set from that column's role,
  not its label.
* **Disposition parameters are derived, not tabulated.** Mo 2024 reports the
  intravenous fit as `A`, `alpha`, `B`, `beta`. The packaged `vc`, `cl`, `q` and
  `vp` are the standard transformation of the mean biexponential and regenerate
  it to rounding, but they are not byte-for-byte quantities from the paper. They
  differ from the separately tabulated `Cl`, `Vss` and peripheral volume by
  -1.7%, -5.0% and -7.0% respectively, because those are means of six
  individually derived values while the packaged set transforms the mean
  biexponential.
* **The oral rate constants are derived from half-lives.** Table 9 reports
  `t1/2ka` and `t1/2ke` rather than `ka` and `ke`; the packaged values are
  `ln(2)` divided by those half-lives. `V/F` is likewise `Cl_F/ke`, since Table 9
  reports no volume.
* **No between-subject variability and no residual error.** Mo 2024 fitted each
  pig separately in Phoenix WinNonlin (WNL5 Classic Modeling) and reported only
  the mean and SD of the individual estimates. A naive two-stage SD of that kind
  confounds true between-animal variability with per-animal estimation
  uncertainty, so it has not been reinterpreted as an IIV variance. Every
  parameter is `fixed()` at the published mean and every residual SD is
  `fixed(0)`; both models simulate deterministically.
* **Emax units follow Table 11, not Figure 5.** Table 11 expresses every Emax in
  ug/mL, while Figure 5 uses a different axis unit for two of the five panels:
  panel B (C4) is labelled ng/mL and panel C (IgG) is labelled g/L; panels A, D
  and E (C3, IgA, IgM) are ug/mL. The packaged values are the Table 11 ug/mL
  numbers, so `dC4` is returned in ug/mL (0.064 ug/mL = 64 ng/mL) and `dIgG` in
  ug/mL (9801.86 ug/mL = 9.80 g/L). Read against its own axis, each Figure 5
  panel confirms the Table 11 unit: C4 rises from roughly 205 to 290 ng/mL, a
  change of about 85 ng/mL against the tabulated Emax of 64 ng/mL, and IgG from
  roughly 12 to 28 g/L, a change of about 16 g/L against the tabulated 9.80 g/L.
  Without this reading the C4 Emax of 0.064 ug/mL looks like a typographical
  error rather than a correctly converted value.
* **Emax sits below the mean peak change visible in Figure 5.** Across all five
  markers the tabulated Emax is roughly 60-75% of the change from the 0 h bar to
  the 0.5 h bar in Figure 5 (approximate values read off the figure, which is not
  tabulated). Because the intravenous C_max is 10 to 15 times each EC50, the
  packaged model predicts essentially the full Emax at 0.5 h, so it
  correspondingly underpredicts the mean peak change in that figure. This is a
  property of the published parameters, not of the packaging: Table 11 reports
  the mean of six individually fitted Emax values whereas Figure 5 plots the
  mean measured concentration, and the two are not constrained to agree. No
  parameter has been adjusted to close the gap.
* **The model cannot reproduce the 0-to-0.5 h rise in Figure 5.** Mo 2024's eq 2
  is a direct-effect model with no effect compartment, driven by an intravenous
  bolus whose concentration is maximal at `t = 0`. The predicted change is
  therefore already near E_max immediately post-dose, whereas the paper's 0 h
  sample is the predose baseline (change zero by definition) and its first
  post-dose sample is at 0.5 h. This is a property of the published model
  structure, not of the packaging; Mo 2024's own conclusion that there is no
  hysteresis between concentration and effect is what justifies the direct-effect
  form. The validation above therefore tests the post-dose samples only.
* **The PD layer models the change from predose, not the absolute level.** Mo
  2024's eq 2 defines `E` as "the absolute changes in complement and
  immunoglobulins". Baseline C3, C4, IgG, IgA and IgM concentrations are shown
  only graphically in Figures 4 and 5 and are not tabulated, so no baseline is
  carried and the models return the increment. Adding a baseline would require
  digitising the figures.
* **No PD model for the intragastric route.** Mo 2024 fitted sigmoid-Emax models
  to the intravenous group only, because the intragastric
  effect-versus-concentration curves were not monotonic (Section 4, Figures S4
  and S5). `Mo_2024_shikimicAcid_pig_oral` is therefore PK-only, and the
  intragastric PD data are represented in the paper only by Pearson correlations
  (Figure 6), which are not a structural model.
* **The double-peak profile is not reproduced.** Two of the six pigs showed a
  second peak at 3 h, which Mo 2024 attributes to enterohepatic recirculation,
  delayed gastric emptying or variable absorption. The single-depot
  first-order model the authors selected does not describe it, and neither does
  the packaged model.
* **Supplement not available.** The Supporting Information (Tables S1-S2,
  Figures S1-S5) is not part of the open-access record on disk. It contains body
  weights, the AIC model-selection table, and diagnostic trend plots -- no
  parameter values used by either model. The body-weight gap is immaterial here
  because all parameters are already per kg.
* **Species.** These are growing pigs, not humans. Both `description` fields and
  the `population$species` metadata are prefixed accordingly.
