---
title: "Everolimus (Mo 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Everolimus (Mo 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Mo_2025_everolimus"))$reference`
* Description: `r rxode2::rxode(readModelDb("Mo_2025_everolimus"))$description`
* Article (open access): <https://doi.org/10.12793/tcp.2025.33.e14>

# Population

Twenty-six healthy adult Korean males were enrolled in a randomized,
open-label, single-dose, 2x2 crossover study at Severance Hospital, Yonsei
University Health System (Seoul, Korea). Participants were 19-55 years old
by protocol (cohort mean 31.0 years, SD 10.8), weighed at least 55 kg
(cohort mean 67.3 kg, SD 8.7), were 171.8 cm tall on average (SD 6.6), and
had a body mass index between 18.5 and 27.0 kg/m^2 (cohort mean 22.7, SD
2.3). All were male (Mo 2025 Table 1).

Each participant received a single 5 mg dose of the reference everolimus
tablet (Afinitor, Novartis Korea) and a single 5 mg dose of the newly
developed dispersible everolimus tablet (SVG-101, test) in the two periods
of the crossover, separated by a washout of at least 10 days. Thirteen
participants were randomized to the R-T sequence and thirteen to the T-R
sequence. Blood was sampled pre-dose and at 0.25, 0.5, 0.75, 1, 1.5, 2,
2.5, 3, 4, 6, 8, 10, 24, 32, 48, 72, 96, 120, and 144 h post-dose.

Only the **test** (dispersible) occasion contributed to the population PK
model: Mo 2025 Methods "Dataset construction for PK analysis" states that
"the data pertaining to the R treatment were excised from the dataset
because our objective was to develop a PK model for the T treatment."
The model therefore used 842 samples from all 26 subjects, while the
non-compartmental analysis in Tables 2 and 3 is based on the 25 subjects
who completed both periods. Below-quantification records were left-censored
via censoring/limit columns rather than discarded. Estimation was performed
in Monolix 2023R1.

The same information is available programmatically via
`readModelDb("Mo_2025_everolimus")()$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Mo_2025_everolimus.R`. The
table below collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `d/dt(depot)` = -ka*depot | n/a | Mo 2025 p. 161, first displayed equation (dG/dt); Fig. 1 schematic |
| `d/dt(central)` | n/a | Mo 2025 p. 161, second displayed equation (dAc/dt) |
| `d/dt(peripheral1)` | n/a | Mo 2025 p. 161, third displayed equation (dAp/dt) |
| Parallel dual absorption (dose split) | n/a | Mo 2025 Fig. 1 caption; Abstract; Results "Model development" |
| `lka` | 9.48 1/h | Table 4, ka, SE 2.99, %RSE 31.6 |
| `lcl` | 7.11 L/h | Table 4, CL, SE 0.40, %RSE 5.6 |
| `lvc` | 47.66 L | Table 4, Vc, SE 2.61, %RSE 5.5 |
| `lq` | 14.32 L/h | Table 4, Q, SE 0.90, %RSE 6.3 |
| `lvp` | 205.94 L | Table 4, Vp, SE 10.89, %RSE 5.3 |
| `ltk0` | 1.12 h | Table 4, Tk0, SE 0.14, %RSE 12.9 (duration; see Errata 1) |
| `ltlag` | 0.63 h | Table 4, Tlag, SE 0.15, %RSE 24.2 |
| `logitfr_zo` | logit(0.62) | Table 4, F, SE 0.05, %RSE 8.0 (logit scale; see Errata 2) |
| `etalka` | 0.94^2 = 0.8836 | Table 4, omega_ka, C.V. 118.88% |
| `etalcl` | 0.28^2 = 0.0784 | Table 4, omega_CL, C.V. 28.76% |
| `etalvc` | 0.26^2 = 0.0676 | Table 4, omega_Vc, C.V. 26.66% |
| `etalq` | 0.31^2 = 0.0961 | Table 4, omega_Q, C.V. 31.30% |
| `etalvp` | 0.26^2 = 0.0676 | Table 4, omega_Vp, C.V. 26.38% |
| `etaltk0` | 0.48^2 = 0.2304 | Table 4, omega_Tk0, C.V. 50.82% |
| `etaltlag` | 0.78^2 = 0.6084 | Table 4, omega_Tlag, C.V. 91.25% |
| `etalogitfr_zo` | 0.76^2 = 0.5776 | Table 4, omega_F, C.V. 27.05% |
| `propSd` | 0.1 | Table 4, sigma_prop, SE 0.0056, %RSE 5.3 |
| Covariate screen (none retained) | n/a | Methods "Population pharmacokinetic modeling"; Results "Model development" |

## Proving the random-effect scale

Mo 2025 Table 4 prints `omega` alongside a `C.V. (%)` column, which lets the
scale of every random effect be verified row by row rather than assumed. For
a log-normal random effect, `CV = sqrt(exp(omega^2) - 1)`.

```{r omega-scale}
omega_tab <- tibble::tribble(
  ~parameter, ~omega, ~cv_published,
  "ka",   0.94, 118.88,
  "CL",   0.28,  28.76,
  "Vc",   0.26,  26.66,
  "Vp",   0.26,  26.38,
  "Q",    0.31,  31.30,
  "Tk0",  0.48,  50.82,
  "Tlag", 0.78,  91.25,
  "F",    0.76,  27.05
) |>
  mutate(cv_lognormal = 100 * sqrt(exp(omega^2) - 1))

omega_tab |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  rename(
    "Parameter"            = parameter,
    "omega (Table 4)"      = omega,
    "C.V. % (Table 4)"     = cv_published,
    "C.V. % if log-normal" = cv_lognormal
  ) |>
  knitr::kable(caption = "Round-tripping every Table 4 omega through the log-normal CV identity.")
```

Seven of the eight rows agree to within rounding of the two-decimal `omega`,
which confirms that Table 4 reports `omega` as the standard deviation of the
random effect on its own transformed scale (so `ini()` takes `omega^2`).

`F` is the sole exception, and it is a large one: a log-normal `F` with
`omega = 0.76` would have a CV of 88%, but Table 4 prints 27.05%. Because
`F` is a bounded proportion, the natural alternative is a logit-normal
distribution, which is also Monolix's default for such a parameter.

```{r omega-f-discriminator}
set.seed(1)
n_draw <- 2e5
eta    <- rnorm(n_draw, 0, 0.76)

cv <- function(x) 100 * sd(x) / mean(x)

tibble::tibble(
  Distribution = c("log-normal", "logit-normal", "probit-normal"),
  `Simulated C.V. %` = round(c(
    cv(0.62 * exp(eta)),
    cv(plogis(qlogis(0.62) + eta)),
    cv(pnorm(qnorm(0.62) + eta))
  ), 2),
  `Table 4 C.V. %` = 27.05
) |>
  knitr::kable(caption = "Only a logit-normal F reproduces the published C.V. for omega_F.")
```

The logit-normal reading reproduces the published 27.05% almost exactly, so
the model file carries `F` on the logit scale (`logitfr_zo` / `expit()`).
This falsifies the paper's blanket prose claim that "we assumed a lognormal
distribution for all the model parameters" -- see Errata 2.

# Virtual cohort

Original observed data are not publicly available. The simulations below use
a virtual cohort of 200 subjects receiving the single 5 mg dispersible dose
on the published sampling schedule.

Each oral administration is encoded as **two parallel dose records at the
same time**: one into `central` carrying the zero-order arm (with
`rate = -2` to invoke the modelled duration `dur(central) = Tk0`) and one
into `depot` carrying the first-order arm. Both records carry the full
nominal 5 mg; the model's `f()` multipliers split it into the
`Fr_zo` / `(1 - Fr_zo)` fractions, so the pair delivers exactly one dose.
Duplicated `(id, time, evid)` keys are therefore expected and correct here;
the uniqueness guard below includes `cmt`, which is the key that must
actually be distinct.

```{r cohort}
set.seed(20250926)

n_sub    <- 200L
dose_mg  <- 5
obs_grid <- c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 2.5, 3, 4, 6, 8, 10,
              24, 32, 48, 72, 96, 120, 144)

ids <- seq_len(n_sub)

# First-order arm: dose to depot, ordinary bolus record (alag(depot) = Tlag
# delays it inside the model).
dose_first <- data.frame(
  id = ids, time = 0, amt = dose_mg, evid = 1L,
  cmt = "depot", rate = NA_real_
)

# Zero-order arm: dose to central with rate = -2 so rxode2 uses the modelled
# duration dur(central) = Tk0.
dose_zero <- transform(dose_first, cmt = "central", rate = -2)

obs <- expand.grid(id = ids, time = obs_grid,
                   KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE)
obs$amt  <- NA_real_
obs$evid <- 0L
obs$cmt  <- "central"
obs$rate <- NA_real_

events <- rbind(dose_first, dose_zero, obs[, names(dose_first)])
events$treatment <- "SVG-101 5 mg (test)"

# The two dose records legitimately share (id, time, evid); cmt is what
# distinguishes them, so that is the key the guard has to check.
stopifnot(nrow(dplyr::distinct(events[, c("id", "time", "evid", "cmt")])) == nrow(events))
```

# Simulation

```{r simulate}
mod <- readModelDb("Mo_2025_everolimus")
sim <- rxode2::rxSolve(mod, events = events, keep = "treatment",
                       returnType = "data.frame")

stopifnot(length(unique(sim$id)) == n_sub, !anyNA(sim$Cc))
```

# Structural check: which absorption arm carries the lag time?

Mo 2025 reports a single `Tlag` and describes the model as "zero-order
kinetics, **followed by** first-order kinetics **with lag time**" (Abstract;
repeated in the Discussion), which attaches the lag to the first-order arm.
The alternative reading -- lag on the zero-order arm -- is simulated here
for comparison, on typical-value (random-effects-zeroed) profiles.

```{r lag-discriminator}
mod_typ <- rxode2::zeroRe(mod)

# Alternative reading: move the lag from the depot arm to the central arm.
mod_alt <- mod |>
  rxode2::model(alag(depot) <- 0) |>
  rxode2::model(alag(central) <- tlag, append = TRUE) |>
  rxode2::zeroRe()

dense <- c(seq(0, 6, by = 0.01), seq(6.5, 168, by = 0.5))
ev1   <- rbind(
  data.frame(id = 1L, time = 0, amt = dose_mg, evid = 1L, cmt = "depot",   rate = NA_real_),
  data.frame(id = 1L, time = 0, amt = dose_mg, evid = 1L, cmt = "central", rate = -2),
  data.frame(id = 1L, time = dense, amt = NA_real_, evid = 0L, cmt = "central", rate = NA_real_)
)

peak_of <- function(m) {
  s <- rxode2::rxSolve(m, ev1, returnType = "data.frame")
  s[which.max(s$Cc), c("time", "Cc")]
}

pA <- peak_of(mod_typ)
pB <- peak_of(mod_alt)

tibble::tibble(
  Reading = c("A: lag on first-order (depot) arm -- as encoded",
              "B: lag on zero-order (central) arm",
              "Observed (Mo 2025 Table 2, test arm)"),
  `Tmax (h)`      = c(round(pA$time, 2), round(pB$time, 2), 1.50),
  `Cmax (ng/mL)`  = c(round(pA$Cc, 2),   round(pB$Cc, 2),   83.38)
) |>
  knitr::kable(caption = "Typical-value Tmax/Cmax under the two lag-time readings. Observed Cmax is the arithmetic mean of Table 2; observed Tmax is the median.")
```

Both readings straddle the observed median Tmax of 1.50 h (reading A peaks
at the end of the zero-order infusion, `Tk0` = 1.12 h; reading B at
`Tlag + Tk0` = 1.75 h), so Tmax alone does not settle the question. Cmax
does: reading A lands within about 1.5% of the observed arithmetic mean of
83.38 ng/mL, whereas reading B is roughly 15% low. Together with the
unambiguous prose, reading A is adopted in the model file.

# Replicate published figures

```{r figure-3}
# Replicates Figure 3 of Mo 2025: visual predictive check of the final model.
sim |>
  group_by(time) |>
  summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  filter(time > 0) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "firebrick", linewidth = 0.8) +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Everolimus concentration (ng/mL)",
    title = "Figure 3 - simulated VPC, 5 mg dispersible everolimus",
    caption = "Replicates Figure 3 of Mo 2025. Median with 5th-95th percentile band, 200 virtual subjects."
  )
```

The paper's key clinical result is the steady-state trough simulation of
Figure 4, which drives the 3-4 mg daily-dose recommendation. Mo 2025
simulated 1,000 virtual subjects per dose over 200-336 h; this vignette uses
200 per arm over the same window.

```{r figure-4}
# Replicates Figure 4 of Mo 2025: steady-state Ctrough distributions at
# 3, 4, and 5 mg once daily against the 5-15 ng/mL target range.
dose_levels <- c(3, 4, 5)
n_ss        <- 200L
dose_times  <- seq(0, 336 - 24, by = 24)
ss_grid     <- seq(200, 336, by = 1)

make_ss_arm <- function(dose, id_offset) {
  arm_ids <- id_offset + seq_len(n_ss)
  dz <- expand.grid(id = arm_ids, time = dose_times,
                    KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE)
  first_order <- data.frame(dz, amt = dose, evid = 1L, cmt = "depot",   rate = NA_real_)
  zero_order  <- data.frame(dz, amt = dose, evid = 1L, cmt = "central", rate = -2)
  ob <- expand.grid(id = arm_ids, time = ss_grid,
                    KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE)
  ob <- data.frame(ob, amt = NA_real_, evid = 0L, cmt = "central", rate = NA_real_)
  out <- rbind(first_order, zero_order, ob)
  out$dose_group <- paste0(dose, " mg once daily")
  out
}

set.seed(336)
ss_events <- dplyr::bind_rows(
  make_ss_arm(dose_levels[1], id_offset = 0L),
  make_ss_arm(dose_levels[2], id_offset = n_ss),
  make_ss_arm(dose_levels[3], id_offset = 2L * n_ss)
)

ss_sim <- rxode2::rxSolve(mod, events = ss_events, keep = "dose_group",
                          returnType = "data.frame")

ss_sim |>
  filter(time >= 200) |>
  group_by(dose_group, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue", linewidth = 0.7) +
  geom_hline(yintercept = c(5, 15), colour = "firebrick") +
  facet_wrap(~dose_group) +
  labs(
    x = "Time (h)", y = "Everolimus concentration (ng/mL)",
    title = "Figure 4 - steady-state concentrations by once-daily dose",
    caption = "Replicates Figure 4 of Mo 2025. Red lines are the 5-15 ng/mL target trough range."
  )
```

```{r figure-4-troughs}
# The paper's dose conclusion is an answer key: 3 mg in range, 4 mg in range
# but elevated, 5 mg above range. Compare the simulated median trough.
trough_times <- seq(216, 336, by = 24)

trough_tab <- ss_sim |>
  filter(time %in% trough_times) |>
  group_by(dose_group) |>
  summarise(
    `5th percentile`  = round(quantile(Cc, 0.05), 2),
    `Median Ctrough`  = round(quantile(Cc, 0.50), 2),
    `95th percentile` = round(quantile(Cc, 0.95), 2),
    .groups = "drop"
  ) |>
  mutate(`Median within 5-15 ng/mL` =
           ifelse(`Median Ctrough` >= 5 & `Median Ctrough` <= 15, "yes", "no")) |>
  rename("Dose group" = dose_group)

knitr::kable(
  trough_tab,
  caption = "Simulated steady-state trough concentrations (ng/mL) vs. the 5-15 ng/mL target."
)
```

This reproduces the paper's conclusion: the 3 mg and 4 mg median troughs sit
inside the 5-15 ng/mL window (with 4 mg noticeably higher), while the 5 mg
median exceeds it -- exactly the basis for Mo 2025's recommendation of 3-4 mg
daily in this healthy-adult population.

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time = 0 row per subject; pre-dose Cc = 0 is correct for an
# extravascular single dose.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- events |>
  dplyr::filter(evid == 1, cmt == "depot") |>
  dplyr::select(id, time, amt, treatment)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

# CL/F and Vz/F are deliberately NOT requested here. PKNCA would form them
# as dose / AUC with the dose in mg and the AUC in h*ng/mL, which yields a
# number 1000x away from L/h and L. Rather than smuggle a scale factor
# through the NCA object, both are reported in the closed-form section
# below, where the mg-to-ng conversion is written out explicitly.
intervals <- data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  auclast    = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

## Comparison against published NCA

Mo 2025 Table 2 reports arithmetic mean (SD) NCA parameters for the test
formulation in 25 subjects, except Tmax, which is reported as a median.
Table 3 additionally reports the geometric least-squares means for AUCinf
(697.90 h*ng/mL) and Cmax (78.87 ng/mL).

```{r nca_compare}
published <- tibble::tribble(
  ~treatment,             ~cmax, ~tmax, ~auclast, ~aucinf.obs, ~half.life,
  "SVG-101 5 mg (test)",  83.38, 1.50,  659.38,   723.43,      35.96
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by        = "treatment",
  units     = c(cmax = "ng/mL", tmax = "h", auclast = "h*ng/mL",
                aucinf.obs = "h*ng/mL", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated vs. published NCA (Mo 2025 Table 2, test formulation). * differs from reference by >20%."
)
```

Cmax, AUClast, AUCinf, and half-life all agree with the published NCA to
within 10%, with AUClast within 1%.

**Tmax is starred, and the star is an artefact of the sampling grid rather
than a model discrepancy.** The simulated median Tmax is 1.0 h against a
published median of 1.5 h -- these are *adjacent points* on the published
sampling schedule (..., 0.75, 1, 1.5, 2, ...), so the smallest resolvable
disagreement on that grid is already a 33% relative difference. The
typical-value Tmax is 1.12 h (the end of the zero-order infusion), which
falls between the two sampling times, so individual subjects round to
either side depending on their sampled `Tk0`, `Tlag`, and `ka`. The
comparison below shows how little separates the two:

```{r tmax-grid}
tmax_sim <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD == "tmax")

tibble::tibble(
  `Sampling time (h)` = c(0.75, 1.0, 1.5, 2.0, 2.5),
  `Simulated subjects with Tmax here` =
    vapply(c(0.75, 1.0, 1.5, 2.0, 2.5),
           \(t) sum(tmax_sim$PPORRES == t), integer(1))
) |>
  knitr::kable(caption = "Distribution of simulated Tmax across the published sampling grid (200 subjects). The published median is 1.5 h; the simulated median is 1.0 h.")
```

No parameter was adjusted to close this gap.

## Closed-form identities

Two disposition quantities can be checked in closed form, independently of
the absorption model and of the NCA implementation.

```{r closed-form}
p <- rxode2::rxode(readModelDb("Mo_2025_everolimus"))$theta
CL <- exp(p[["lcl"]]); Vc <- exp(p[["lvc"]])
Q  <- exp(p[["lq"]]);  Vp <- exp(p[["lvp"]])

kel <- CL / Vc; k12 <- Q / Vc; k21 <- Q / Vp
b   <- kel + k12 + k21
lambda_z <- (b - sqrt(b^2 - 4 * kel * k21)) / 2

# The 1000 factor converts the 5 mg dose to ng so that Dose / AUC lands in
# L/h against an AUC expressed in h*ng/mL -- the same conversion the model
# applies in `Cc <- 1000 * central / vc`.
tibble::tibble(
  Quantity = c("AUCinf for 5 mg (h*ng/mL)", "Terminal half-life (h)",
               "CL/F (L/h)", "Vz = CL / lambda_z (L)", "Vss = Vc + Vp (L)"),
  Model = round(c(1000 * 5 / CL, log(2) / lambda_z, CL, CL / lambda_z, Vc + Vp), 2),
  Published = c(697.90, 35.96, 7.44, 378.44, NA),
  Source = c("Table 3 GeoLSM AUCinf", "Table 2 t1/2", "Table 2 CL/F",
             "Table 2 Vd/F", "not reported")
) |>
  knitr::kable(caption = "Closed-form disposition identities vs. published NCA.")
```

`AUCinf = Dose / CL` is exact for a linear model and matches the published
geometric least-squares mean to within 1%, which simultaneously gates the
clearance estimate, the mg-to-ng unit conversion in `Cc`, and the fact that
the two parallel dose records deliver exactly one dose rather than two.

# Assumptions and deviations

- **Whole blood as the observation matrix.** `compartmentData` records
  `central` and `peripheral1` as whole blood. Mo 2025 uses the boilerplate
  phrase "plasma drug concentration-time curve" in its NCA parameter
  definitions, but four lines of evidence point to whole blood: samples were
  drawn into EDTA-K2 tubes and aliquoted with no centrifugation step
  described (Methods "Study design"); the 5-15 ng/mL target trough range the
  paper simulates against is the standard whole-blood everolimus
  therapeutic-drug-monitoring range; the observed concentration magnitudes
  are whole-blood-scale (everolimus partitions strongly into erythrocytes,
  so plasma concentrations run several-fold lower); and everolimus
  monitoring is conventionally whole blood.
- **Cohort composition.** The virtual cohort carries no covariates because
  the final model has none. Age, weight, height, and BMI were screened by
  the authors and none was retained; they are recorded in the model file's
  `covariatesDataExcluded` for provenance only.
- **Cohort size.** Figures 3 and 4 use 200 virtual subjects per arm rather
  than the 1,000 per dose used by Mo 2025 for Figure 4. The published
  median and percentile bands are stable at this size; the reduction keeps
  the vignette inside the package's render-time budget.
- **Tmax comparison.** The simulated Tmax is read off the published discrete
  sampling grid, so it is quantized to the same grid the observed median was
  read from. The typical-value Tmax (1.12 h) falls between the 1 h and 1.5 h
  sampling times.
- **No between-occasion variability.** The crossover design would in
  principle support inter-occasion variability, but only the test occasion
  entered the model, and Mo 2025 Table 4 reports a single IIV term per
  parameter with no IOV component.

## Errata and deviations from the published text

1. **`Tk0` is a duration, not a rate constant.** Mo 2025 Table 4 labels
   `Tk0` the "Zero-order absorption rate constant", but three things
   contradict that label and agree with each other: the Table 4 row header
   gives its unit as `(h)` rather than `1/h`; the Figure 1 caption defines
   `Tk0` explicitly as "zero-order absorption duration (h)"; and the symbol
   `Tk0` is Monolix's standard name for the duration of a zero-order
   absorption process. The model file uses the duration reading
   (`dur(central) <- tk0`). Under the rate-constant reading the printed unit
   would be wrong and the absorption phase would last about 0.9 h with no
   consistent parameterization, so the Table 4 description is treated as the
   error.
2. **The zero-order fraction `F` is logit-normal, not log-normal.** Mo 2025
   Methods states "We assumed a lognormal distribution for all the model
   parameters", but Table 4's own `C.V. (%)` column falsifies that for `F`:
   `omega_F = 0.76` yields a log-normal CV of 88.4% against the printed
   27.05%, whereas a logit-normal `F` about `logit(0.62)` yields 26.8%
   (a probit-normal yields 39.9%). Every other row round-trips correctly
   under the log-normal identity. The model file therefore carries `F` on
   the logit scale, which also keeps simulated values inside `(0, 1)` as a
   proportion requires. See the "Proving the random-effect scale" section
   above for the row-by-row check.
3. **The lag time attaches to the first-order arm.** Mo 2025 reports one
   `Tlag` without stating which absorption arm it delays. The Abstract and
   Discussion both phrase the model as "zero-order kinetics, followed by
   first-order kinetics with lag time", and the Cmax check in the
   "Structural check" section above supports that reading over the
   alternative. The model file uses `alag(depot) <- tlag`.
4. **Model n = 26 vs. NCA n = 25.** The population PK model was fit to all
   26 enrolled subjects, while Tables 2 and 3 report NCA on the 25 who
   completed both crossover periods. One subject withdrew after day 4 but
   contributed test-formulation data to the model. Comparisons in this
   vignette are therefore between a 26-subject model and 25-subject NCA
   summaries.
5. **Table 3 Cmax point-estimate typo.** Mo 2025 Table 3 prints the Cmax
   geometric mean ratio as "0.6832 (0.6832-0.8604)" -- the point estimate
   erroneously repeats the lower confidence limit. Two independent
   recomputations give the same corrected value: the GeoLSM columns of the
   same table yield 78.87 / 102.87 = 0.7667, and the geometric centre of the
   printed interval is sqrt(0.6832 * 0.8604) = 0.7667. The point estimate
   should therefore read 0.7667, with the interval (0.6832-0.8604) intact.
   This value is not used anywhere in the model and is recorded here only
   for completeness.
