---
title: "Search models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Search models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
```

Type any combination of drug, author, year or keyword. Every word must match
(case-insensitive), and each word is looked for in the drug names, model names,
first author, year, full citation and description. Narrow a word to one field
with a prefix:

| Example                | Finds                                                   |
|------------------------|---------------------------------------------------------|
| `abacavir`             | every abacavir model                                    |
| `zhao 2012`            | papers by Zhao from 2012                                |
| `author:Stein`         | papers with Stein anywhere in the author list           |
| `drug:adalimumab`      | only the drug and model names are searched              |
| `year:2020-2024 mab`   | 2020-2024 papers mentioning "mab"                       |
| `pediatric vancomycin` | vancomycin models whose description mentions pediatrics |

The search is shareable: the page address updates to `?q=<your search>` as you
type. The full table is also on the [List of models](list-of-models.html) page.

```{r build-index, include = FALSE}
db <- nlmixr2lib::modeldb

# Locate the model files: the installed package when building the site, the
# source tree when knitting this file directly from vignettes/articles/.
modelDir <- system.file("modeldb", package = "nlmixr2lib")
if (!nzchar(modelDir) || !dir.exists(modelDir)) {
  modelDir <- file.path("..", "..", "inst", "modeldb")
}

# Pull the `reference <- ...` string out of a model file without calling the
# model function. Most files build it with paste() over string literals, so the
# right-hand side is evaluated in baseenv(), where nothing but base functions
# can be reached.
getReference <- function(f) {
  if (!file.exists(f)) {
    return(NA_character_)
  }
  exprs <- tryCatch(parse(f, keep.source = FALSE), error = function(e) NULL)
  isAssign <- function(x) {
    is.call(x) && as.character(x[[1]]) %in% c("<-", "=")
  }
  for (e in exprs) {
    if (!isAssign(e) || !is.call(e[[3]]) ||
          !identical(e[[3]][[1]], as.name("function"))) {
      next
    }
    for (s in as.list(e[[3]][[3]])[-1]) {
      if (isAssign(s) && identical(s[[2]], as.name("reference"))) {
        v <- tryCatch(eval(s[[3]], baseenv()), error = function(e) NULL)
        if (is.character(v)) {
          return(paste(v, collapse = " "))
        }
      }
    }
  }
  NA_character_
}
db$reference <- vapply(
  file.path(modelDir, db$filename), getReference, character(1),
  USE.NAMES = FALSE
)

# Author and year come from the `Author_Year_drug` filename convention; files
# that do not follow it (the generic PK/PD templates) leave both blank. The
# drug is the part of the label before " (Author Year)".
parts <- strsplit(tools::file_path_sans_ext(basename(db$filename)), "_",
                  fixed = TRUE)
hasAuthorYear <- vapply(parts, function(p) {
  length(p) >= 3 && grepl("^[0-9]{4}[a-z]?$", p[2]) && p[1] != "NA"
}, logical(1))
db$author <- ifelse(hasAuthorYear, vapply(parts, `[`, "", 1), "")
db$year <- ifelse(hasAuthorYear, vapply(parts, `[`, "", 2), "")
db$drug <- sub(" \\([^()]*\\)$", "", db$label)

# One search record per vignette: a vignette often validates several models
# (several drugs, species or model variants from the same paper). Models
# without a vignette are kept as records of their own, without a link.
db$key <- ifelse(is.na(db$vignette), paste0("model:", db$name), db$vignette)
db <- db[order(tolower(db$label)), , drop = FALSE]
records <- lapply(split(db, factor(db$key, levels = unique(db$key))),
                  function(g) {
  first <- function(x) {
    x <- x[!is.na(x) & nzchar(x)]
    if (length(x) == 0) "" else x[[1]]
  }
  list(
    v = first(g$vignette),
    c = first(g$category),
    a = first(g$author),
    y = first(g$year),
    d = unique(g$drug),
    n = g$name,
    r = first(g$reference),
    s = first(g$description)
  )
})
names(records) <- NULL

# Minimal JSON writer (strings and string arrays only), so the page needs no
# extra package. `</` is escaped so no string can close the <script> element.
jsonString <- function(x) {
  x <- gsub("\\", "\\\\", x, fixed = TRUE)
  x <- gsub('"', '\\"', x, fixed = TRUE)
  x <- gsub("\n", "\\n", x, fixed = TRUE)
  x <- gsub("\r", "\\r", x, fixed = TRUE)
  x <- gsub("\t", "\\t", x, fixed = TRUE)
  x <- gsub("[[:cntrl:]]", " ", x)
  x <- gsub("</", "<\\/", x, fixed = TRUE)
  paste0('"', x, '"')
}
jsonRecord <- function(r) {
  fields <- vapply(names(r), function(k) {
    val <- if (length(r[[k]]) == 1 && !k %in% c("d", "n")) {
      jsonString(r[[k]])
    } else {
      paste0("[", paste(jsonString(r[[k]]), collapse = ","), "]")
    }
    paste0('"', k, '":', val)
  }, character(1))
  paste0("{", paste(fields, collapse = ","), "}")
}
indexJson <- paste0(
  "[", paste(vapply(records, jsonRecord, character(1)), collapse = ",\n"), "]"
)
stopifnot(length(records) > 0)
```

```{r search-ui, echo = FALSE, results = "asis"}
# Everything below is emitted as a pandoc RAW block so pandoc does not reparse
# the HTML, CSS or JavaScript as markdown (see list-of-models.Rmd for how that
# goes wrong on `$` and `_`).
ui <- r"--(
<style>
#model-search-controls { display: flex; flex-wrap: wrap; gap: 0.5em; margin: 1em 0; }
#model-search-q { flex: 1 1 20em; }
#model-search-controls select, #model-search-controls input[type=number] { width: auto; }
#model-search-results td { vertical-align: top; }
#model-search-results td.ms-desc { font-size: 0.9em; }
#model-search-results .ms-models { font-size: 0.8em; color: #6c757d; }
</style>
<div id="model-search-controls">
  <input id="model-search-q" class="form-control" type="search"
         placeholder="Search by drug, author, year or keyword, e.g. abacavir zhao 2012"
         aria-label="Search models" autocomplete="off">
  <select id="model-search-cat" class="form-select" aria-label="Model group">
    <option value="">All groups</option>
    <option value="specificDrugs" selected>Specific drugs</option>
    <option value="ddmore">DDMoRe</option>
    <option value="other">General</option>
  </select>
  <input id="model-search-from" class="form-control" type="number" min="1900" max="2100"
         placeholder="Year from" aria-label="Year from">
  <input id="model-search-to" class="form-control" type="number" min="1900" max="2100"
         placeholder="Year to" aria-label="Year to">
</div>
<p id="model-search-count" aria-live="polite"></p>
<table id="model-search-results" class="table table-sm table-striped">
  <thead><tr><th>Drug(s)</th><th>Author</th><th>Year</th><th>Description</th></tr></thead>
  <tbody></tbody>
</table>
<p><button id="model-search-more" class="btn btn-outline-secondary btn-sm" type="button" hidden>Show all results</button></p>
<script type="application/json" id="model-index">
@@INDEX@@
</script>
<script>
(function () {
  var data = JSON.parse(document.getElementById("model-index").textContent);
  var LIMIT = 200;
  var q = document.getElementById("model-search-q");
  var cat = document.getElementById("model-search-cat");
  var yFrom = document.getElementById("model-search-from");
  var yTo = document.getElementById("model-search-to");
  var count = document.getElementById("model-search-count");
  var tbody = document.querySelector("#model-search-results tbody");
  var more = document.getElementById("model-search-more");
  var showAll = false;

  data.forEach(function (r) {
    r.yn = parseInt(r.y, 10);
    r.fAuthor = (r.a + " " + r.r).toLowerCase();
    r.fDrug = (r.d.join(" ") + " " + r.n.join(" ")).toLowerCase();
    r.fAll = [r.d.join(" "), r.n.join(" "), r.a, r.y, r.r, r.s, r.v]
      .join(" ").toLowerCase();
  });

  function esc(s) {
    return String(s).replace(/[&<>"]/g, function (c) {
      return { "&": "&amp;", "<": "&lt;", ">": "&gt;", '"': "&quot;" }[c];
    });
  }

  // Each whitespace-separated term must match. `field:value` restricts a term
  // to one field; `year:` also takes a range such as 2015-2020.
  function parseQuery(text) {
    return text.toLowerCase().split(/\s+/).filter(Boolean).map(function (t) {
      var m = t.match(/^(author|drug|year):(.*)$/);
      return m ? { field: m[1], value: m[2] } : { field: null, value: t };
    }).filter(function (t) { return t.value.length > 0; });
  }

  function matchTerm(r, t) {
    if (t.field === "year") {
      var range = t.value.match(/^(\d{4})?-(\d{4})?$/);
      if (range) {
        var lo = range[1] ? +range[1] : -Infinity;
        var hi = range[2] ? +range[2] : Infinity;
        return r.yn >= lo && r.yn <= hi;
      }
      return r.y.toLowerCase().indexOf(t.value) === 0;
    }
    if (t.field === "author") return r.fAuthor.indexOf(t.value) >= 0;
    if (t.field === "drug") return r.fDrug.indexOf(t.value) >= 0;
    return r.fAll.indexOf(t.value) >= 0;
  }

  function render() {
    var terms = parseQuery(q.value);
    var c = cat.value;
    var lo = yFrom.value ? +yFrom.value : null;
    var hi = yTo.value ? +yTo.value : null;
    var hits = data.filter(function (r) {
      if (c && r.c !== c) return false;
      if (lo !== null && !(r.yn >= lo)) return false;
      if (hi !== null && !(r.yn <= hi)) return false;
      return terms.every(function (t) { return matchTerm(r, t); });
    });
    var shown = showAll ? hits : hits.slice(0, LIMIT);
    tbody.innerHTML = shown.map(function (r) {
      var drugs = esc(r.d.join(", "));
      var title = r.v ? '<a href="' + esc(r.v) + '.html">' + drugs + "</a>" : drugs;
      var desc = r.s.length > 240 ? r.s.slice(0, 237) + "..." : r.s;
      return "<tr><td>" + title +
        '<div class="ms-models">' + esc(r.n.join(", ")) + "</div></td>" +
        "<td>" + esc(r.a) + "</td><td>" + esc(r.y) + "</td>" +
        '<td class="ms-desc" title="' + esc(r.r) + '">' + esc(desc) + "</td></tr>";
    }).join("");
    count.textContent = hits.length + " of " + data.length + " entries match" +
      (shown.length < hits.length ? " (showing the first " + shown.length + ")" : "") + ".";
    more.hidden = shown.length >= hits.length;
  }

  function syncUrl() {
    var url = new URL(window.location.href);
    if (q.value) url.searchParams.set("q", q.value); else url.searchParams.delete("q");
    window.history.replaceState(null, "", url);
  }

  var timer = null;
  function onInput() {
    clearTimeout(timer);
    timer = setTimeout(function () { showAll = false; render(); syncUrl(); }, 150);
  }

  q.value = new URLSearchParams(window.location.search).get("q") || "";
  [q, yFrom, yTo].forEach(function (el) { el.addEventListener("input", onInput); });
  cat.addEventListener("change", onInput);
  more.addEventListener("click", function () { showAll = true; render(); });
  render();
  q.focus();
})();
</script>
)--"
# sub() with fixed = TRUE so backslashes in the JSON are inserted verbatim.
knitr::raw_block(sub("@@INDEX@@", indexJson, ui, fixed = TRUE), "html")
```
