---
title: "Tocilizumab in ICU-admitted COVID-19 (Moes 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tocilizumab in ICU-admitted COVID-19 (Moes 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Moes_2022_tocilizumab"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Moes_2022_tocilizumab"))$description`
* Article: [Clin Pharmacokinet. 2022;61(2):231-247](https://doi.org/10.1007/s40262-021-01074-2)
  (open access, PMC8502793). Electronic supplementary material (ESM) 1 contains
  the final NONMEM control stream, which this extraction uses to settle the
  units of Vmax and the IIV block.

# Population

Moes 2022 is a single-centre, open-label, observational PK study at the Leiden
University Medical Center (December 2020 to March 2021). It enrolled 29 adults
with PCR-confirmed COVID-19 who were admitted to the ICU with respiratory organ
support (82.8% mechanically ventilated). Every patient received dexamethasone
6 mg once daily for up to 10 days and a single intravenous tocilizumab dose of
8 mg/kg (maximum 800 mg) within 24 h of starting organ support.

Baseline characteristics (Moes 2022 Table 1): mean age 64 years (45-80),
72.4% male, mean body weight 96.4 kg (58-130), mean BMI 30.9 kg/m^2
(20.1-46.1), mean albumin 32 g/L, and mean CRP 146 mg/L. The administered doses
ranged from 472 to 1552 mg (mean 781 mg). One patient received a double dose by
accident. A total of 139 free-tocilizumab plasma concentrations (ELISA, LLOQ
0.2 ug/mL) were collected from day 1 to day 20 after the dose, with 1-11
samples per patient.

The same information is available programmatically via
`readModelDb("Moes_2022_tocilizumab")$population`.

# Source trace

The final model is a one-compartment model with parallel first-order and
Michaelis-Menten elimination (Moes 2022 Fig. 1). No covariate was retained:
body weight entered as a power function on CL gave an exponent of 0.002
(Section 3.5).

| Quantity | Value | Source |
|---|---|---|
| Structure: `d/dt(central) = -(CL/V) * central - Vmax * C / (Km + C)` | -- | Fig. 1; ESM 1 `$DES` |
| CL (L/day) | 0.725 | Table 3; ESM 1 `$THETA` |
| Vd (L) | 4.34 | Table 3; ESM 1 `$THETA` |
| Vmax (mg/day) | 4.19 | Table 3; ESM 1 `$THETA` (amount rate in `$DES`) |
| Km (mg/L = ug/mL) | 0.22 | Table 3; ESM 1 `$THETA` |
| omega^2 CL | 0.0351 (18.9% CV) | ESM 1 `$OMEGA BLOCK(2)`; CV in Table 3 |
| cov(CL, Vd) | 0.0355 (r = 0.914) | ESM 1 `$OMEGA BLOCK(2)` (not printed in Table 3) |
| omega^2 Vd | 0.043 (21.0% CV) | ESM 1 `$OMEGA BLOCK(2)`; CV in Table 3 |
| IIV on Vmax, Km | none | Table 2; ESM 1 `$OMEGA 0 FIX` |
| Proportional residual error | 0.171 | Table 3; ESM 1 `$THETA` |
| Additive residual error (ug/mL) | 0.139 | Table 3; ESM 1 `$THETA` |
| Residual-error form | combined2, `sqrt(prop^2 * IPRED^2 + add^2)` | ESM 1 `$ERROR` |

The stored variances reproduce the Table 3 CVs as `sqrt(exp(omega^2) - 1)`:

```{r iiv-check}
mod <- readModelDb("Moes_2022_tocilizumab")
ui <- rxode2::rxode2(mod)
om <- ui$omega
cv <- 100 * sqrt(exp(diag(om)) - 1)
cv
stopifnot(
  abs(cv[["etalcl"]] - 18.9) < 0.05,
  abs(cv[["etalvc"]] - 21.0) < 0.05
)
```

# Structural checks

## The Michaelis-Menten arm is active

The model has variables named `cl` and `vc`. The check below confirms that
rxode2 integrates the explicit ODE, including the saturable arm, rather than
replacing it with an analytical linear solution. The typical AUC0-inf must
drop below `Dose/CL` when Vmax is non-zero.

```{r mm-active}
stopifnot(is.null(ui$linCmt))
mod_typ <- rxode2::zeroRe(ui)
dose_mean <- 21879 / 29 # mg, mean first dose (ESM Table S3 total / 29 patients)
ev_typ <- rxode2::et(amt = dose_mean, cmt = "central", rate = dose_mean * 24) |>
  rxode2::et(c(seq(0, 1, by = 0.01), seq(1.05, 60, by = 0.05)))
trap <- function(x, y) sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)
s_mm <- rxode2::rxSolve(mod_typ, ev_typ, returnType = "data.frame")
s_lin <- rxode2::rxSolve(
  rxode2::ini(mod_typ, lvmax = log(1e-12)),
  ev_typ,
  returnType = "data.frame"
)
auc_typ <- c(
  with_mm = trap(s_mm$time, s_mm$Cc),
  linear_only = trap(s_lin$time, s_lin$Cc),
  dose_over_cl = dose_mean / 0.725
)
auc_typ
stopifnot(
  # linear-only solve equals the closed form Dose/CL
  abs(auc_typ[["linear_only"]] / auc_typ[["dose_over_cl"]] - 1) < 0.005,
  # the saturable arm removes roughly 10% of the exposure at this dose
  auc_typ[["with_mm"]] < 0.95 * auc_typ[["linear_only"]]
)
```

## Linear versus non-linear clearance (Figure 6)

The non-linear clearance at concentration `C` is `Vmax / (Km + C)`. It
exceeds the linear CL below `C* = Vmax / CL - Km`. Moes 2022 reports this
crossover as approximately 5 ug/mL (Section 3.4, Fig. 6). The value follows
from Table 3 alone, and it holds only if Vmax is an amount rate (mg/day). If
Vmax were a concentration rate (mg/L/day), as in some other tocilizumab
models, the crossover would be `Vmax * V / CL - Km` = 24.9 ug/mL.

```{r fig6}
p <- c(cl = 0.725, v = 4.34, vmax = 4.19, km = 0.22)
c_star <- p[["vmax"]] / p[["cl"]] - p[["km"]]
c_star
stopifnot(abs(c_star - 5.56) < 0.01)

fig6 <- tibble(conc = 10^seq(-2, 2.5, length.out = 300)) |>
  mutate(
    cl_nonlin = p[["vmax"]] / (p[["km"]] + conc),
    cl_total = p[["cl"]] + cl_nonlin,
    `Linear CL` = p[["cl"]] / cl_total,
    `Non-linear CL` = cl_nonlin / cl_total
  ) |>
  pivot_longer(c(`Linear CL`, `Non-linear CL`), names_to = "pathway", values_to = "fraction")

ggplot(fig6, aes(conc, fraction, colour = pathway)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = c_star, linetype = "dashed") +
  scale_x_log10() +
  labs(
    x = "Tocilizumab concentration (ug/mL)",
    y = "Fraction of total clearance",
    colour = NULL,
    caption = "Replicates Figure 6 of Moes 2022; dashed line = crossover at 5.56 ug/mL."
  ) +
  theme_bw()
```

# Virtual cohort

The observed cohort is re-used directly. ESM Table S3 lists the dose
prescribed to each of the 29 patients (472-800 mg, mean 754 mg). Table 1
reports a mean administered dose of 781 mg with a maximum of 1552 mg. That
summary is reproduced exactly when the patient prescribed 776 mg received
it twice, which is the accidental double dose described in Section 3.1:
(21879 + 776) / 29 = 781.2 mg. The published exposure metric is the AUC0-inf
of the *first* dose, so the virtual cohort gives each patient their Table S3
dose once. Each of the 29 doses is replicated 6 times, giving 174 virtual
patients. Each dose is given as a 1-h intravenous infusion.

```{r cohort}
dose_s3 <- c(
  760, 800, 640, 664, 640, 799, 472, 800, 776, 640, 800, 688, 800, 800, 800,
  800, 800, 600, 800, 800, 800, 800, 800, 800, 800, 800, 800, 800, 800
)
stopifnot(length(dose_s3) == 29L, sum(dose_s3) == 21879)
# Table 1 mean and maximum administered dose, with patient 9's dose given twice
dose_admin <- dose_s3
dose_admin[9] <- 2 * dose_s3[9]
stopifnot(abs(mean(dose_admin) - 781) < 0.5, max(dose_admin) == 1552)

n_rep <- 6L
cohort <- tibble(
  id = seq_len(29L * n_rep),
  dose = rep(dose_s3, times = n_rep)
)
obs_times <- c(seq(0, 1, by = 0.02), seq(1.1, 10, by = 0.1), seq(10.25, 40, by = 0.25))
```

# Simulation

```{r simulate-cohort}
rxode2::rxSetSeed(20211011)
ev_cohort <- bind_rows(
  cohort |> mutate(time = 0, amt = dose, rate = dose * 24, evid = 1L, cmt = "central"),
  tidyr::crossing(cohort, time = obs_times) |>
    mutate(amt = 0, rate = 0, evid = 0L, cmt = "central")
) |>
  arrange(id, time, desc(evid)) |>
  select(id, time, amt, rate, evid, cmt, dose)

sim_cohort <- rxode2::rxSolve(
  mod, ev_cohort,
  keep = "dose", returnType = "data.frame"
) |>
  mutate(treatment = "Observed-cohort doses")
```

The individual predictions (`ipredSim`, no residual error) are plotted,
matching the individual-prediction curves in Figure 3 of Moes 2022.

```{r fig3, fig.height = 4.5}
sim_cohort |>
  filter(id <= 29) |>
  ggplot(aes(time, ipredSim, group = id)) +
  geom_line(alpha = 0.6) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_log10(limits = c(0.1, NA)) +
  coord_cartesian(xlim = c(0, 25)) +
  labs(
    x = "Time after dose (days)",
    y = "Tocilizumab (ug/mL)",
    caption = "Compare with the individual-prediction curves of Figure 3 of Moes 2022 (29 of 174 virtual patients shown)."
  ) +
  theme_bw()
```

# PKNCA validation

```{r pknca}
# After the saturable arm clears the drug, the ODE solver returns values of
# order -1e-11 ug/mL (integration round-off around zero). Confirm that is all
# they are, then set them to zero so PKNCA's log-down trapezoid is defined.
stopifnot(min(sim_cohort$ipredSim) > -1e-8)
conc_df <- sim_cohort |>
  filter(!is.na(ipredSim)) |>
  transmute(id, time, conc = pmax(ipredSim, 0), treatment)
dose_df <- cohort |>
  transmute(id, time = 0, dose, treatment = "Observed-cohort doses")

conc_obj <- PKNCA::PKNCAconc(conc_df, conc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, dose ~ time | treatment + id)
intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, clast.obs = TRUE, auclast = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_wide <- as.data.frame(nca_res) |>
  select(id, treatment, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
summary(nca_wide[, c("cmax", "clast.obs", "auclast")])
```

Below about 1 ug/mL the saturable pathway dominates elimination. The
concentration then falls so quickly (effective half-life about 0.2 day) that by
day 40 it is numerically zero in every virtual patient. A log-linear terminal
fit is not meaningful on that tail, so AUC0-40 d (`auclast`) is used as
AUC0-inf. The first assertion below checks that the tail is negligible.

Moes 2022 reports the mean (SD) first-dose AUC0-inf across the 29 patients as
938 (190) ug/mL*day (Section 3.4; ESM Tables S1 and S2). The same tables report
an "average concentration" of 44.0 ug/mL. The ESM 1 control stream defines this
quantity as `CAV1 = (DOSE/CL)/24`, i.e. the linear-only `Dose/CL` averaged over
24 days.

```{r nca-compare}
stopifnot(
  !anyNA(nca_wide$auclast),
  max(nca_wide$clast.obs) < 1e-3
)
published <- data.frame(treatment = "Observed-cohort doses", auclast = 938)
sim_mean <- nca_wide |>
  group_by(treatment) |>
  summarise(auclast = mean(auclast), .groups = "drop")

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_mean,
  reference = published,
  by = "treatment",
  units = c(auclast = "ug*day/mL"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = "Cohort-mean AUC0-40 d (equal to AUC0-inf; concentrations are zero by day 40) of the simulated cohort vs the published mean AUC0-inf of Moes 2022 (938 ug/mL*day). * marks rows differing by >20%."
)

# Cavg as defined in the control stream: mean over subjects of (Dose/CL_i)/24.
cl_i <- sim_cohort |> distinct(id, cl, dose)
cavg_sim <- mean(cl_i$dose / cl_i$cl) / 24
c(cavg_sim = cavg_sim, cavg_paper = 44.0)

stopifnot(
  # Centre of the distribution: a mis-transcribed CL, Vmax or dose unit
  # would move the mean AUC by tens of percent.
  abs(mean(nca_wide$auclast) / 938 - 1) < 0.10,
  abs(cavg_sim / 44.0 - 1) < 0.10,
  # Deterministic typical value at the mean dose (no random draws).
  abs(auc_typ[["with_mm"]] / 938 - 1) < 0.03
)
```

The simulated SD of AUC0-inf across the virtual patients is
`r round(sd(nca_wide$auclast))` ug/mL*day, compared with the published 190.
The published SD comes from individual (post hoc) estimates of 29 patients. The
simulated SD comes from 174 random draws from the population distribution.

# Dosing scenarios (Figures 7-9)

The helper below simulates one arm of 200 virtual patients (the per-arm cap)
and summarises the median and 90% prediction interval. It also returns, for
each patient, the last time the concentration is above 1 and 5 ug/mL.

```{r scenario-helper}
sim_arm <- function(regimen, label, n = 200L, t_end = 40, seed = 1) {
  rxode2::rxSetSeed(seed)
  doses <- regimen |> tidyr::crossing(id = seq_len(n))
  obs <- tidyr::crossing(
    id = seq_len(n),
    time = c(seq(0, 1, by = 0.05), seq(1.1, t_end, by = 0.1))
  )
  ev <- bind_rows(
    doses |> mutate(rate = amt * 24, evid = 1L, cmt = "central"),
    obs |> mutate(amt = 0, rate = 0, evid = 0L, cmt = "central")
  ) |>
    arrange(id, time, desc(evid)) |>
    select(id, time, amt, rate, evid, cmt)
  rxode2::rxSolve(mod, ev, returnType = "data.frame") |>
    mutate(treatment = label)
}

time_above <- function(sim, thr) {
  sim |>
    group_by(treatment, id) |>
    summarise(t_above = max(c(0, time[ipredSim > thr])), .groups = "drop") |>
    mutate(threshold = thr)
}

pi_summary <- function(sim) {
  sim |>
    group_by(treatment, time) |>
    summarise(
      med = median(ipredSim),
      lo = quantile(ipredSim, 0.05),
      hi = quantile(ipredSim, 0.95),
      .groups = "drop"
    )
}

plot_pi <- function(summ, caption) {
  ggplot(summ, aes(time, med)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey70", alpha = 0.6) +
    geom_line() +
    geom_hline(yintercept = c(1, 5), linetype = "dashed") +
    facet_wrap(~treatment) +
    scale_y_log10(limits = c(0.1, NA)) +
    labs(x = "Time (days)", y = "Tocilizumab (ug/mL)", caption = caption) +
    theme_bw()
}
```

## 8 mg/kg versus 600 mg fixed dose (Figure 7)

Figure 7 simulated patients of 60, 70, 80, 90 and 100 kg receiving 8 mg/kg,
compared with 600 mg for all. Because the model has no body-weight covariate,
each 8 mg/kg arm is simply a fixed dose of 480-800 mg.

```{r fig7, fig.height = 5.5}
fig7_arms <- tibble(
  label = c(paste0("8 mg/kg, ", c(60, 70, 80, 90, 100), " kg"), "600 mg fixed"),
  amt = c(8 * c(60, 70, 80, 90, 100), 600)
)
sim7 <- bind_rows(lapply(seq_len(nrow(fig7_arms)), function(i) {
  sim_arm(
    tibble(time = 0, amt = fig7_arms$amt[i]),
    fig7_arms$label[i],
    seed = 700 + i
  )
})) |>
  mutate(treatment = factor(treatment, levels = fig7_arms$label))
plot_pi(pi_summary(sim7), "Replicates Figure 7 of Moes 2022 (median and 90% PI; 200 per arm).")
```

## Alternative fixed doses (Figure 8)

```{r fig8, fig.height = 5}
fig8_doses <- c(800, 600, 400, 200)
sim8 <- bind_rows(lapply(seq_along(fig8_doses), function(i) {
  sim_arm(
    tibble(time = 0, amt = fig8_doses[i]),
    paste(fig8_doses[i], "mg"),
    seed = 800 + i
  )
})) |>
  mutate(treatment = factor(treatment, levels = paste(fig8_doses, "mg")))
plot_pi(pi_summary(sim8), "Replicates Figure 8 of Moes 2022 (median and 90% PI; 200 per arm).")

ta8 <- bind_rows(time_above(sim8, 1), time_above(sim8, 5))
ta8_tbl <- ta8 |>
  group_by(treatment, threshold) |>
  summarise(
    median_days = median(t_above),
    p05_days = quantile(t_above, 0.05),
    .groups = "drop"
  )
ta8_tbl |>
  dplyr::rename(
    "Regimen" = treatment,
    "Threshold (ug/mL)" = threshold,
    "Median days above" = median_days,
    "5th percentile days above" = p05_days
  ) |>
  knitr::kable(digits = 1)
```

Moes 2022 reports that the 600 mg fixed dose keeps the concentration above
both thresholds for about 15 days or longer, and that even 200 mg keeps it
above them for at least 7 days (Section 3.5). The gate below checks the median
and the 5th percentile, not the minimum of a random draw.

```{r fig8-gate}
get_ta <- function(tbl, trt, thr, col) {
  tbl[[col]][tbl$treatment == trt & tbl$threshold == thr]
}
stopifnot(
  get_ta(ta8_tbl, "600 mg", 5, "median_days") > 14,
  get_ta(ta8_tbl, "200 mg", 5, "median_days") > 7,
  get_ta(ta8_tbl, "200 mg", 1, "p05_days") > 7
)
```

## Split dosing with the same cumulative dose (Figure 9)

```{r fig9, fig.height = 4}
sim9 <- bind_rows(
  sim_arm(tibble(time = 0, amt = 800), "800 mg once", seed = 901),
  sim_arm(tibble(time = c(0, 10), amt = 400), "400 mg q10d x2", seed = 902),
  sim_arm(tibble(time = c(0, 7, 14, 21), amt = 200), "200 mg q7d x4", seed = 903)
) |>
  mutate(treatment = factor(treatment, levels = c("800 mg once", "400 mg q10d x2", "200 mg q7d x4")))
plot_pi(pi_summary(sim9), "Replicates Figure 9 of Moes 2022 (median and 90% PI; 200 per arm).")

ta9 <- time_above(sim9, 1) |>
  group_by(treatment) |>
  summarise(median_days_above_1 = median(t_above), .groups = "drop")
knitr::kable(ta9, digits = 1)
```

Moes 2022 concludes that repeated lower doses keep the concentration above the
saturation thresholds for longer than one 800 mg dose (Section 3.6). The
typical-value solve below checks this without random draws.

```{r fig9-gate}
typ_above <- function(regimen) {
  ev <- rxode2::et(amt = regimen$amt[1], cmt = "central", rate = regimen$amt[1] * 24)
  for (k in seq_len(nrow(regimen))[-1]) {
    ev <- rxode2::et(ev, time = regimen$time[k], amt = regimen$amt[k], cmt = "central", rate = regimen$amt[k] * 24)
  }
  ev <- rxode2::et(ev, seq(0, 60, by = 0.05))
  s <- rxode2::rxSolve(mod_typ, ev, returnType = "data.frame")
  max(s$time[s$Cc > 1])
}
typ9 <- c(
  once_800 = typ_above(tibble(time = 0, amt = 800)),
  q10d_400 = typ_above(tibble(time = c(0, 10), amt = 400)),
  q7d_200 = typ_above(tibble(time = c(0, 7, 14, 21), amt = 200))
)
typ9
stopifnot(typ9[["q10d_400"]] > typ9[["once_800"]], typ9[["q7d_200"]] > typ9[["q10d_400"]])
```

# Assumptions and deviations

* **IIV block.** Table 3 reports only the CVs of the CL and Vd random effects.
  The final control stream in ESM 1 estimates them as a `BLOCK(2)` with a
  covariance of 0.0355 (correlation 0.914), and its variances reproduce the
  Table 3 CVs exactly. The packaged model uses that block.
* **Vmax units.** The abstract prints Vmax as "4.19 ug/day". The Results
  text, Table 3, and the ESM 1 `$DES` block (where Vmax is subtracted from an
  amount rate) all give mg/day, and the 5 ug/mL crossover in Fig. 6 holds only
  for mg/day. The abstract unit is treated as a typographical error.
* **Infusion duration.** The paper does not state the infusion duration. All
  simulations use a 1-h intravenous infusion, as in the tocilizumab label. The
  exposure summaries used here are insensitive to this choice.
* **Accidental double dose.** The prescribed doses in ESM Table S3 average
  754 mg. Table 1 reports a mean administered dose of 781 mg with a maximum of
  1552 mg. These reconcile exactly if the patient prescribed 776 mg received
  that dose twice. The published AUC is of the first dose, so the virtual
  cohort gives every patient their Table S3 dose once. The patient whose dose
  was split into two infusions within 12 h is modelled as one infusion. When
  the double dose was instead simulated as one 1552 mg infusion, the cohort-mean
  AUC was 5% above the published value and the SD was 315 ug/mL*day instead
  of 190. With the first-dose convention both match closely.
* **AUC summary statistic.** The published 938 (190) ug/mL*day is a mean of
  individual (post hoc) AUCs. It is compared here with the mean over the
  simulated cohort and with the deterministic typical value at the mean dose.
* **Pharmacodynamics.** CRP and sIL-6R were described only descriptively
  (Figures 3-5). No PK/PD model was estimated, so none is packaged.
* **Screened covariates.** Body weight, age, sex, BMI, BSA, height, and the
  laboratory covariates were screened and not retained. They are listed in the
  model's `covariatesDataExcluded` metadata for provenance.
* **Errata.** A literature check (Europe PMC, 2026-09-29) found no published
  erratum or correction; the only linked record is the preprint.
