---
title: "Tacrolimus IR-Tac to LCP-Tac conversion (Mohammed Ali 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tacrolimus IR-Tac to LCP-Tac conversion (Mohammed Ali 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
mod <- rxode2::rxode(readModelDb("MohammedAli_2025_tacrolimus"))
```

* Citation: `r mod$reference`
* Article: <https://doi.org/10.3390/pharmaceutics17091185>
* Supplement (Figure S1 NPDE, Table S1 simulations):
  <https://www.mdpi.com/article/10.3390/pharmaceutics17091185/s1>

Mohammed Ali 2025 is the first population PK analysis built specifically to
answer a *conversion* question: when a stable renal transplant recipient is
switched from twice-daily immediate-release tacrolimus (IR-Tac, Prograf) to
once-daily extended-release LCP-Tac (Envarsus, MeltDose), what dose ratio
reproduces the previous exposure -- and does the answer depend on *CYP3A5*
genotype? European guidelines recommend a single 1:0.7 ratio regardless of
genetics. The paper's simulations conclude that 1:0.7 is right for *CYP3A5*
non-expressers but that expressers need 1:0.6.

The structural model is a two-compartment model with delayed first-order
absorption, fitted jointly to both formulations in the *same* 30 patients. Three
features make it unusual and are the focus of the validation below:

1. **Formulation-specific absorption.** Separate absorption rate constants
   (`ka` 2.04 vs 0.111 h^-1) and lag times (0.465 vs 1.4 h) for IR-Tac and
   LCP-Tac. The MeltDose product absorbs roughly 18-fold more slowly.
2. **A 24 h circadian rhythm** on apparent clearance and, for IR-Tac only, on
   its absorption rate constant. This is what produces the day-dose /
   night-dose asymmetry visible in the paper's Figure 2 panels (a) and (b).
3. **Genotype-by-formulation bioavailability.** *CYP3A5* enters the model on
   `F`, not on `CL/F`, giving four relative bioavailabilities anchored at
   `F = 1` for the LCP-Tac non-expresser reference group.

# Population

`r mod$population$n_subjects` stable adult renal transplant recipients
(`r mod$population$regions`), all at least six months post-transplant and on
tacrolimus + mycophenolate mofetil + prednisone, enrolled in the open-label
single-centre trial NCT02961608. Mean body weight was 72 kg (IQR 64-80), mean
age 58 years (IQR 48-68), and 8 of 30 (27%) were female (Table 1). Renal
function was typical for a stable graft (eGFR 49.6 mL/min, IQR 34-57) and
hematocrit was near-normal (40.9%, IQR 37.6-44.8). Genotyping of *CYP3A5\*3*
(rs776746) found 9 *\*1/\*3*, 1 *\*1/\*1* and 20 *\*3/\*3* patients; because a
single *\*1/\*1* patient cannot support its own stratum, the analysis pooled
*\*1* carriers into 10 **expressers** and kept 20 **non-expressers**.

Each patient contributed two rich 24 h steady-state profiles -- one a week
before conversion on IR-Tac (481 samples) and one four weeks after conversion
on LCP-Tac (451 samples), 932 whole-blood concentrations in total, with 10-18
samples per profile. This within-patient crossover is what lets the model
separate a formulation effect on `F` from between-subject differences.

The same information is available programmatically:

```{r population}
str(readModelDb("MohammedAli_2025_tacrolimus")()$population, max.level = 1)
```

# Source trace

Every value below is also recorded as an in-file comment next to its `ini()`
entry in `inst/modeldb/specificDrugs/MohammedAli_2025_tacrolimus.R`.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL/F mesor) | 11.9 L/h | Table 3, "CL/F" (RSE 8.5%) |
| `lvc` (Vc/F) | 78 L | Table 3, "Vc/F" (RSE 14.7%) |
| `lq` (CLd/F) | 25.8 L/h | Table 3, "CLd/F" (RSE 8.5%) |
| `lvp` (Vp/F) | 500 L, **FIX** | Table 3, "Vp/F = 500 FIX"; Results 3.2.1 |
| `lka_ir` | 2.04 /h | Table 3, "Ka IR-Tac" (RSE 40%) |
| `lka_lcp` | 0.111 /h | Table 3, "Ka LCP_Tac" (RSE 16.9%) |
| `ltlag_ir` | 0.465 h | Table 3, "Lag-Time IR-Tac" (RSE 0.1%) |
| `ltlag_lcp` | 1.4 h | Table 3, "Lag-Time LCP-Tac" (RSE 2.4%) |
| `lfdepot_lcp_nonexpr` | 1, **FIX** | Table 3, "F LCP-Tac_PM = 1 FIX"; Methods 2.6.1 eq. 2 |
| `lfdepot_ir_nonexpr` | 0.745 | Table 3, "F IR-Tac_PM" (RSE 7.6%) |
| `lfdepot_lcp_expr` | 0.693 | Table 3, "F LCP-Tac_HM" (RSE 13.7%) |
| `lfdepot_ir_expr` | 0.427 | Table 3, "F IR-Tac_HM" (RSE 13.4%) |
| `acrophase_cl` | 17 h | Table 3, "Acrophase CL/F" (RSE 3.6%) |
| `amp_cl` | 3.42 L/h | Table 3, "Amp CL/F" (RSE 17.1%) |
| `acrophase_ka_ir` | 3.13 h | Table 3, "Acrophase ka" (RSE 18.3%) |
| `amp_ka_ir` | 1.55 /h | Table 3, "Amp ka" (RSE 44.5%) |
| `propSd` | 0.133 | Table 3, "RE. (-) = 13.30" (RSE 8.2%); Results 3.2.1 |
| `etalcl` | 26.49 %CV | Table 3, "IIV CL/F" (RSE 29.1%) |
| `etalvc` | 53.47 %CV | Table 3, "IIV Vc/F" (RSE 42%) |
| `etalka_ir` | 150.66 %CV | Table 3, "IIV Ka IR-Tac" (RSE 25.6%) |
| `etalka_lcp` | 67.23 %CV | Table 3, "IIV Ka LCP_Tac" (RSE 46.5%) |
| block corr. Vc/F--Ka IR | 0.7563 | Table 3, "Vc/F/Ka IR-Tac Correlation" |
| block corr. Vc/F--Ka LCP | 0.4438 | Table 3, "Vc/F/Ka LCP-Tac Correlation" |
| block corr. Ka IR--Ka LCP | 0.45 | Table 3, "Ka IR-Tac/Ka LCP-Tac Correlation" |
| `etaiov_cl_*` | 20.85 %CV | Table 3, "IOV CL" (RSE 23.9%) |
| `etaiov_vc_*` | 58.82 %CV | Table 3, "IOV Vc" (RSE 28.9%) |
| Cosinor equation `P = theta1 + theta_AMP * cos(2*pi*(TIME - ACROPHASE)/1440)` | n/a | Methods 2.6.1, equation 3 |
| Relative bioavailability `F = 1 * theta_x` | n/a | Methods 2.6.1, equation 2 |
| Two-compartment ODEs, delayed first-order absorption | n/a | Results 3.2.1 |

Reference values used as validation targets: Table 2 (observed NCA by
formulation and genotype), the Discussion's four derived `CL/F` values, and
supplementary Table S1 (the paper's own 1000-subject simulations of AUC24 and
Ctrough across IR-Tac doses 0.5-10 mg and conversion ratios 0.5-1.0).

# Simulation setup

The model's time variable is **clock time in hours from midnight**, because the
acrophases (17:00 for `CL/F`, 03:08 for the IR-Tac `ka`) are absolute times of
day rather than times after dose. The paper defines `TIME` as "the time in
minutes starting at midnight of the first PK profile" but never states the
clock time at which the simulated doses were given, so the helper below takes it
as an argument; `dose_hour = 0` (the literal reading, in which the first dose
sits at the `TIME` origin) is used throughout, and the sensitivity to this
choice is quantified in a later section.

```{r helpers}
`%||%` <- function(a, b) if (is.null(a)) b else a

# One arm = one dosing regimen for one genotype. Built as a plain data frame:
# assigning covariate columns onto an rxEt object silently drops them.
# `obs_by` defaults to a fine 0.02 h grid: the IR-Tac peak is sharp (tmax ~1.3 h)
# and a coarse grid costs several tenths of a percent of AUC, which is more than
# the closed-form gate below tolerates. Use a coarser grid for large cohorts.
#
# `n_days = 30` is set by the terminal half-life, which is 45.8 h for these
# parameters (eigenvalues of the two-compartment system with Vp/F = 500 L).
# 30 days is 15.7 terminal half-lives and reproduces the closed-form AUC to
# 0.002%; 14 days would still be 0.53% short of steady state. Note that the
# paper simulated 'after at least 10 days of treatment' (Methods 2.8), which
# is only 5.2 terminal half-lives, so its Table S1 values sit slightly below
# true steady state.
make_arm <- function(daily_mg, form_lcp, expr, n_sub = 1L, id_offset = 0L,
                     dose_hour = 0, n_days = 30L, obs_by = 0.02, label = NULL) {
  tau <- if (form_lcp == 1) 24 else 12
  amt <- if (form_lcp == 1) daily_mg else daily_mg / 2
  # doses from the origin through the end of the observation window
  n_dose <- as.integer(n_days * 24 / tau) + 1L
  dose_times <- dose_hour + seq(0, by = tau, length.out = n_dose)
  # observe the final full 24 h day, which is the steady-state window
  win_start <- (n_days - 1L) * 24
  obs_times <- seq(win_start, win_start + 24, by = obs_by)

  ids <- id_offset + seq_len(n_sub)
  dose_rows <- expand.grid(id = ids, time = dose_times) |>
    mutate(amt = amt, evid = 1L, cmt = "depot")
  obs_rows <- expand.grid(id = ids, time = obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")

  bind_rows(dose_rows, obs_rows) |>
    mutate(
      CYP3A5_EXPR       = expr,
      FORM_TAC_ENVARSUS = form_lcp,
      OCC               = if (form_lcp == 1) 2L else 1L,
      daily_mg          = daily_mg,
      genotype          = if (expr == 1) "CYP3A5*1 expresser" else "CYP3A5 non-expresser",
      formulation       = if (form_lcp == 1) "LCP-Tac" else "IR-Tac",
      treatment         = label %||% paste0(formulation, " ", daily_mg, " mg/day"),
      win_start         = win_start
    ) |>
    arrange(id, time, desc(evid))
}

# Typical-value solve (IIV and residual error switched off).
mod_typ <- rxode2::zeroRe(mod)

solve_typ <- function(ev, model = mod_typ) {
  keep <- c("CYP3A5_EXPR", "FORM_TAC_ENVARSUS", "daily_mg", "genotype",
            "formulation", "treatment", "win_start")
  out <- rxode2::rxSolve(model, ev, keep = keep, returnType = "data.frame",
                         atol = 1e-10, rtol = 1e-10, maxsteps = 1e6)
  # rxSolve omits `id` for a single-subject event table (failure pattern 8)
  if (is.null(out$id)) out$id <- 1L
  out
}

# Exact AUC over [a, b] by trapezoid on the dense solution grid.
auc_dense <- function(time, conc, a, b) {
  k <- time >= a - 1e-9 & time <= b + 1e-9
  tt <- time[k]; cc <- conc[k]
  sum(diff(tt) * (head(cc, -1) + tail(cc, -1)) / 2)
}
```

# Gate 1: the closed-form steady-state AUC identity

With the circadian amplitudes set to zero the model is time-invariant, so at
steady state the AUC over one full day must equal `daily dose * F / (CL/F)`
*exactly*. This single identity gates the ODE system, the dose encoding, the
bioavailability wiring and the observation scaling all at once, so it is worth
asserting before anything else.

```{r gate-closed-form}
F_tab <- tibble::tribble(
  ~formulation, ~expr, ~F_rel,
  "LCP-Tac",    0,     1.000,   # Table 3: F LCP-Tac_PM = 1 FIX
  "IR-Tac",     0,     0.745,   # Table 3: F IR-Tac_PM
  "LCP-Tac",    1,     0.693,   # Table 3: F LCP-Tac_HM
  "IR-Tac",     1,     0.427    # Table 3: F IR-Tac_HM
)
CL_MESOR <- 11.9

# amplitudes zeroed -> time-invariant model
mod_flat <- mod_typ |> rxode2::ini(amp_cl = 0, amp_ka_ir = 0)

gate1 <- F_tab |>
  rowwise() |>
  mutate(
    daily_mg = 4,
    sim = {
      ev <- make_arm(daily_mg, form_lcp = as.integer(formulation == "LCP-Tac"),
                     expr = expr)
      s <- solve_typ(ev, mod_flat)
      auc_dense(s$time, s$Cc, s$win_start[1], s$win_start[1] + 24)
    },
    closed_form = daily_mg * F_rel / CL_MESOR * 1000
  ) |>
  ungroup() |>
  mutate(pct_diff = 100 * (sim - closed_form) / closed_form)

gate1 |>
  mutate(genotype = ifelse(expr == 1, "expresser", "non-expresser")) |>
  select(formulation, genotype, F_rel, closed_form, sim, pct_diff) |>
  rename("Formulation" = formulation, "CYP3A5" = genotype,
         "F (rel.)" = F_rel, "dose*F/CL (ng*h/mL)" = closed_form,
         "Simulated AUC24 (ng*h/mL)" = sim, "% diff" = pct_diff) |>
  knitr::kable(digits = 3, caption =
    "Time-invariant model: simulated steady-state AUC24 vs the closed form.")

stopifnot(nrow(gate1) == 4L, max(abs(gate1$pct_diff)) < 0.05)
```

The identity holds to better than 0.05% in all four formulation-by-genotype
groups, which gates the ODE system, the dose encoding, the bioavailability
wiring and the `ng/mL` scaling together.

# Gate 2: the Discussion's four derived CL/F values

Because `F` is a pure multiplier on dose, the apparent clearance of each group
is `11.9 / F`. The Discussion states these as 11.9, 15.97, 17.17 and 27.87 L/h
for LCP-Tac non-expressers, IR-Tac non-expressers, LCP-Tac expressers and
IR-Tac expressers respectively. This is an independent arithmetic check on the
bioavailability parameterization: if `F` had been coded as an effect on
clearance instead of on dose, these would not reproduce.

```{r gate-derived-cl}
published_clf <- c("LCP-Tac 0" = 11.90, "IR-Tac 0" = 15.97,
                   "LCP-Tac 1" = 17.17, "IR-Tac 1" = 27.87)

gate2 <- F_tab |>
  mutate(
    key       = paste(formulation, expr),
    derived   = CL_MESOR / F_rel,
    published = unname(published_clf[key])
  )
stopifnot(!anyNA(gate2$published))          # fail loudly on a label mismatch

gate2 |>
  mutate(genotype = ifelse(expr == 1, "expresser", "non-expresser"),
         `% diff` = 100 * (derived - published) / published) |>
  select(formulation, genotype, F_rel, published, derived, `% diff`) |>
  rename("Formulation" = formulation, "CYP3A5" = genotype, "F (rel.)" = F_rel,
         "Published CL/F (L/h)" = published, "11.9 / F (L/h)" = derived) |>
  knitr::kable(digits = c(0, 0, 3, 2, 2, 3), caption =
    "Discussion-quoted apparent clearances reproduced as 11.9 / F.")

stopifnot(max(abs(gate2$derived - gate2$published)) < 0.01)
```

All four agree to within 0.01 L/h, i.e. to the precision at which the paper
prints them.

# Gate 3: does the circadian rhythm on CL/F act on LCP-Tac?

The paper is explicit that the `ka` rhythm belongs to IR-Tac ("the absorption
rate constant of the IR-Tac formulation"), but it does not say whether the
`CL/F` rhythm applies to LCP-Tac records as well. `CL/F` is a single shared
parameter of a joint model, and a diurnal metabolic rhythm is a property of the
patient rather than of the tablet, so the model applies it to both formulations.

Table S1 settles the question quantitatively. Integrating a cosinor-modulated
clearance over a **full** 24 h cycle gives an AUC slightly *above* the
time-invariant `dose * F / CL` -- a Jensen effect, because AUC depends on
`1 / CL`. For a once-daily formulation this excess is independent of the dose
clock time, so it is a clean discriminator.

```{r gate-circadian-scope}
lcp_targets <- tibble::tribble(          # Table S1, LCP-Tac columns
  ~daily_ir, ~expr, ~ratio, ~tableS1,
  0.5, 1, 1.0,  29.86,   0.5, 0, 1.0,  43.56,
  3.0, 1, 1.0, 180.73,   3.0, 0, 1.0, 267.27,
  6.0, 1, 1.0, 358.32,   6.0, 0, 1.0, 522.73,
  0.5, 1, 0.6,  18.19,   0.5, 0, 0.6,  25.60,
  3.0, 1, 0.6, 107.98,   3.0, 0, 0.6, 158.18,
  6.0, 1, 0.6, 218.28,   6.0, 0, 0.6, 307.25
)

auc_lcp <- function(daily_mg, expr, model) {
  ev <- make_arm(daily_mg, form_lcp = 1L, expr = expr)
  s <- solve_typ(ev, model)
  auc_dense(s$time, s$Cc, s$win_start[1], s$win_start[1] + 24)
}

gate3 <- lcp_targets |>
  rowwise() |>
  mutate(
    dose_lcp   = daily_ir * ratio,
    with_rhythm = auc_lcp(dose_lcp, expr, mod_typ),
    no_rhythm   = auc_lcp(dose_lcp, expr, mod_flat)
  ) |>
  ungroup() |>
  mutate(
    dev_with = 100 * (with_rhythm - tableS1) / tableS1,
    dev_none = 100 * (no_rhythm   - tableS1) / tableS1
  )

tibble::tibble(
  Reading = c("CL/F rhythm applied to LCP-Tac (model as shipped)",
              "CL/F rhythm suppressed on LCP-Tac"),
  `Mean |% dev| vs Table S1` = c(mean(abs(gate3$dev_with)), mean(abs(gate3$dev_none))),
  `Max |% dev|`              = c(max(abs(gate3$dev_with)),  max(abs(gate3$dev_none)))
) |>
  knitr::kable(digits = 2, caption =
    "Discriminating the scope of the CL/F circadian rhythm against Table S1.")

# The applied reading must be the better one, by a clear margin.
stopifnot(mean(abs(gate3$dev_with)) < 0.5 * mean(abs(gate3$dev_none)),
          max(abs(gate3$dev_with)) < 3)
```

Applying the rhythm to LCP-Tac reduces the mean absolute deviation from the
paper's own simulations by roughly a factor of five, and every one of the 12
cells lands within 3%. Suppressing it leaves a systematic ~3.5% shortfall in
the same direction throughout. The model therefore carries the `CL/F` rhythm on
both formulations.

# Replicating supplementary Table S1

Table S1 reports geometric means of AUC24 and Ctrough from 1000 simulated
subjects for every IR-Tac dose and conversion ratio. For this model the
geometric mean of AUC over a log-normal cohort equals the typical-value AUC
(AUC scales as `1 / CL` and `log CL` is symmetric), so the typical-value
profile is the right comparator and is far cheaper than re-running 1000
subjects per cell.

```{r tableS1-grid}
grid_targets <- tibble::tribble(
  ~daily_ir, ~expr, ~ratio, ~auc_S1, ~ctrough_S1,
  # IR-Tac reference arms (ratio NA); Table S1 "IR-Tac0-24"
  3.0, 1, NA, 105.76, 3.19,   3.0, 0, NA, 183.98, 5.52,
  6.0, 1, NA, 213.39, 6.45,   6.0, 0, NA, 367.08, 10.97,
  # LCP-Tac arms at the two clinically recommended ratios
  3.0, 1, 0.6, 107.98, 3.41,  3.0, 0, 0.6, 158.18, 5.00,
  3.0, 1, 0.7, 126.40, 3.96,  3.0, 0, 0.7, 181.84, 5.75,
  6.0, 1, 0.6, 218.28, 6.90,  6.0, 0, 0.6, 307.25, 9.72,
  6.0, 1, 0.7, 253.90, 8.01,  6.0, 0, 0.7, 369.14, 11.70
)

sim_cell <- function(daily_ir, expr, ratio) {
  is_lcp <- !is.na(ratio)
  dose <- if (is_lcp) daily_ir * ratio else daily_ir
  ev <- make_arm(dose, form_lcp = as.integer(is_lcp), expr = expr)
  s <- solve_typ(ev)
  w0 <- s$win_start[1]
  tibble::tibble(
    auc_sim     = auc_dense(s$time, s$Cc, w0, w0 + 24),
    ctrough_sim = approx(s$time, s$Cc, xout = w0 + 24)$y
  )
}

s1_cmp <- grid_targets |>
  rowwise() |>
  mutate(sim_cell(daily_ir, expr, ratio)) |>
  ungroup() |>
  mutate(
    arm = ifelse(is.na(ratio), "IR-Tac (reference)",
                 paste0("LCP-Tac ratio ", ratio)),
    genotype = ifelse(expr == 1, "expresser", "non-expresser"),
    auc_pct     = 100 * (auc_sim - auc_S1) / auc_S1,
    ctrough_pct = 100 * (ctrough_sim - ctrough_S1) / ctrough_S1
  )

s1_cmp |>
  select(daily_ir, genotype, arm, auc_S1, auc_sim, auc_pct,
         ctrough_S1, ctrough_sim, ctrough_pct) |>
  rename("IR dose (mg/day)" = daily_ir, "CYP3A5" = genotype, "Arm" = arm,
         "AUC24 Table S1" = auc_S1, "AUC24 model" = auc_sim,
         "AUC % diff" = auc_pct, "Ctrough Table S1" = ctrough_S1,
         "Ctrough model" = ctrough_sim, "Ctrough % diff" = ctrough_pct) |>
  knitr::kable(digits = 2, caption =
    "Model typical-value predictions vs the paper's supplementary Table S1 simulations (AUC24 in ng*h/mL, Ctrough in ng/mL).")

stopifnot(nrow(s1_cmp) == 12L,
          max(abs(s1_cmp$auc_pct)) < 4,
          max(abs(s1_cmp$ctrough_pct)) < 7)
```

Agreement is good throughout: LCP-Tac AUC24 is within 1.5% of Table S1 in every
cell, and the IR-Tac arms sit about 3% high. The IR-Tac offset is discussed
under Assumptions and deviations -- it is driven by the one quantity the paper
never reports, the clock time of the simulated doses.

# The conversion-ratio question (Figures 3 and 4)

Figures 3 and 4 of the paper are boxplots of simulated AUC24 and Ctrough for
IR-Tac total daily doses of 3-6 mg against conversion ratios 0.5-1.0, split by
genotype. The panel below reproduces the typical-value backbone of those
figures: the LCP-Tac exposure curve against conversion ratio, with the IR-Tac
reference exposure drawn as a horizontal line. Where the curve crosses the line
is the ratio that preserves exposure -- the paper's central result.

```{r figure34, fig.height = 5.5}
ratios <- seq(0.5, 1.0, by = 0.1)
doses  <- c(3, 4, 5, 6)

lcp_curve <- expand.grid(daily_ir = doses, expr = c(0, 1), ratio = ratios) |>
  rowwise() |>
  mutate(sim_cell(daily_ir, expr, ratio)) |>
  ungroup()

ir_ref <- expand.grid(daily_ir = doses, expr = c(0, 1)) |>
  rowwise() |>
  mutate(sim_cell(daily_ir, expr, NA)) |>
  ungroup() |>
  rename(auc_ir = auc_sim, ctrough_ir = ctrough_sim)

lab_geno <- function(x) ifelse(x == 1, "CYP3A5*1 expresser", "CYP3A5 non-expresser")

ggplot(lcp_curve, aes(ratio, auc_sim, colour = factor(daily_ir))) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  geom_hline(data = ir_ref, aes(yintercept = auc_ir, colour = factor(daily_ir)),
             linetype = "dashed", linewidth = 0.4) +
  facet_wrap(~lab_geno(expr)) +
  labs(x = "IR-Tac to LCP-Tac conversion ratio",
       y = "Steady-state AUC24 (ng*h/mL)",
       colour = "IR-Tac dose\n(mg/day)",
       title = "Replicates Figure 3 of Mohammed Ali 2025 (typical-value backbone)",
       subtitle = "Dashed line: exposure on the pre-conversion IR-Tac regimen") +
  theme_bw()
```

The ratio at which LCP-Tac exposure matches the previous IR-Tac exposure is the
paper's recommendation. Solving for it directly:

```{r optimal-ratio}
opt_ratio <- lcp_curve |>
  left_join(ir_ref |> select(daily_ir, expr, auc_ir, ctrough_ir),
            by = c("daily_ir", "expr")) |>
  group_by(daily_ir, expr) |>
  summarise(
    ratio_auc     = approx(auc_sim, ratio, xout = auc_ir[1])$y,
    ratio_ctrough = approx(ctrough_sim, ratio, xout = ctrough_ir[1])$y,
    .groups = "drop"
  ) |>
  mutate(genotype = lab_geno(expr))

opt_ratio |>
  select(genotype, daily_ir, ratio_auc, ratio_ctrough) |>
  arrange(genotype, daily_ir) |>
  rename("CYP3A5" = genotype, "IR-Tac dose (mg/day)" = daily_ir,
         "Exposure-neutral ratio (AUC24)" = ratio_auc,
         "Exposure-neutral ratio (Ctrough)" = ratio_ctrough) |>
  knitr::kable(digits = 3, caption =
    "Conversion ratio that reproduces the pre-conversion IR-Tac exposure. The paper recommends 1:0.6 for expressers and 1:0.7 for non-expressers.")

# The paper's headline conclusion, asserted: expressers need a LOWER ratio than
# non-expressers, and the two cluster around 0.6 and 0.7 respectively.
by_geno <- opt_ratio |> group_by(expr) |> summarise(r = mean(ratio_auc))
r_expr    <- by_geno$r[by_geno$expr == 1]
r_nonexpr <- by_geno$r[by_geno$expr == 0]
stopifnot(length(r_expr) == 1L, length(r_nonexpr) == 1L)
stopifnot(r_expr < r_nonexpr)                       # genotype ordering
stopifnot(abs(r_expr - 0.6) < 0.05)                 # 1:0.6 for expressers
stopifnot(abs(r_nonexpr - 0.7) < 0.05)              # 1:0.7 for non-expressers
```

Both recommendations fall out of the model: expressers need roughly a 0.6 ratio
(a 40% dose reduction) and non-expressers roughly 0.7 (a 30% reduction), and the
answer is essentially independent of the starting dose because the model is
linear in dose. This reproduces the paper's central conclusion, and it is a
consequence of the `F` estimates alone: the expresser LCP-Tac/IR-Tac
bioavailability ratio is `0.693 / 0.427 = 1.62`, versus
`1.000 / 0.745 = 1.34` in non-expressers, and `1 / 1.62 = 0.617` while
`1 / 1.34 = 0.745`.

# Day-dose versus night-dose asymmetry (Figure 2)

Panels (a) and (b) of Figure 2 show the IR-Tac day dose and night dose
separately, which only makes sense because the circadian terms make them
different. Table S1 quantifies it: AUC over 0-12 h exceeds AUC over 12-24 h by
about 17% in every dose-genotype cell.

```{r figure2-asymmetry}
ir_profile <- make_arm(3, form_lcp = 0L, expr = 1L, obs_by = 0.02) |> solve_typ()
w0 <- ir_profile$win_start[1]

a_day   <- auc_dense(ir_profile$time, ir_profile$Cc, w0,      w0 + 12)
a_night <- auc_dense(ir_profile$time, ir_profile$Cc, w0 + 12, w0 + 24)

tibble::tibble(
  Interval = c("0-12 h (day dose)", "12-24 h (night dose)", "ratio day/night"),
  Model    = c(a_day, a_night, a_day / a_night),
  `Table S1 (3 mg, expresser)` = c(56.78, 48.53, 56.78 / 48.53)
) |>
  knitr::kable(digits = 3, caption =
    "IR-Tac day-dose vs night-dose steady-state AUC (ng*h/mL).")

# The asymmetry must be present and in the direction the paper reports.
stopifnot(a_day > a_night, a_day / a_night > 1.05)
```

```{r figure2-profiles, fig.height = 4}
prof <- bind_rows(
  make_arm(3, form_lcp = 0L, expr = 1L, obs_by = 0.02) |> solve_typ(),
  make_arm(3, form_lcp = 0L, expr = 0L, obs_by = 0.02) |> solve_typ(),
  make_arm(3 * 0.6, form_lcp = 1L, expr = 1L, obs_by = 0.02) |> solve_typ(),
  make_arm(3 * 0.7, form_lcp = 1L, expr = 0L, obs_by = 0.02) |> solve_typ()
) |>
  mutate(clock = (time - win_start) %% 24)

ggplot(prof, aes(clock, Cc, colour = formulation)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~genotype) +
  scale_x_continuous(breaks = seq(0, 24, by = 6)) +
  labs(x = "Clock time within the steady-state day (h from midnight)",
       y = "Whole-blood tacrolimus (ng/mL)", colour = "Formulation",
       title = "Steady-state typical-value profiles at exposure-neutral doses",
       subtitle = "IR-Tac 3 mg/day split q12h; LCP-Tac at the genotype-specific conversion ratio") +
  theme_bw()
```

The IR-Tac profile shows the sharp twice-daily peaks the formulation is known
for, with visibly unequal day and night peaks, while LCP-Tac (`ka` = 0.111
h^-1) produces the flat once-daily profile that motivated the MeltDose product.

# Virtual cohort and PKNCA validation

The stochastic cohort below carries the published IIV, the correlated
Vc/F--ka block and the inter-occasion variability. It is used for the
NCA validation and for a variability band around the typical-value profiles.

```{r cohort}
set.seed(20250912)
N_PER_ARM <- 100L   # cap is 200 per arm

cohort_ev <- bind_rows(
  make_arm(3,       form_lcp = 0L, expr = 1L, n_sub = N_PER_ARM, id_offset =   0L, obs_by = 0.25,
           label = "IR-Tac 3 mg/day, expresser"),
  make_arm(3,       form_lcp = 0L, expr = 0L, n_sub = N_PER_ARM, id_offset = 100L, obs_by = 0.25,
           label = "IR-Tac 3 mg/day, non-expresser"),
  make_arm(3 * 0.6, form_lcp = 1L, expr = 1L, n_sub = N_PER_ARM, id_offset = 200L, obs_by = 0.25,
           label = "LCP-Tac 1.8 mg/day, expresser"),
  make_arm(3 * 0.7, form_lcp = 1L, expr = 0L, n_sub = N_PER_ARM, id_offset = 300L, obs_by = 0.25,
           label = "LCP-Tac 2.1 mg/day, non-expresser")
)

sim <- rxode2::rxSolve(
  mod, cohort_ev,
  keep = c("CYP3A5_EXPR", "FORM_TAC_ENVARSUS", "daily_mg", "genotype",
           "formulation", "treatment", "win_start"),
  returnType = "data.frame", atol = 1e-8, rtol = 1e-8, maxsteps = 1e6
)
if (is.null(sim$id)) sim$id <- 1L
stopifnot(dplyr::n_distinct(sim$id) == 4L * N_PER_ARM, all(sim$Cc >= 0))
```

```{r cohort-plot, fig.height = 4.5}
bands <- sim |>
  mutate(clock = (time - win_start) %% 24) |>
  group_by(treatment, formulation, genotype, clock) |>
  summarise(p05 = quantile(Cc, 0.05), p50 = median(Cc),
            p95 = quantile(Cc, 0.95), .groups = "drop")

ggplot(bands, aes(clock, p50, colour = formulation, fill = formulation)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.18, colour = NA) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~genotype) +
  scale_x_continuous(breaks = seq(0, 24, by = 6)) +
  labs(x = "Clock time within the steady-state day (h from midnight)",
       y = "Whole-blood tacrolimus (ng/mL)",
       colour = "Formulation", fill = "Formulation",
       title = "Simulated steady-state variability (median, 5th-95th percentiles)",
       subtitle = paste0(N_PER_ARM, " subjects per arm; compare Figure 2 of Mohammed Ali 2025")) +
  theme_bw()
```

Non-compartmental analysis over the steady-state dosing day, with `PKNCA`
configured to use the linear-up / log-down trapezoid that the paper's own NCA
used (Methods 2.5).

```{r pknca}
conc_df <- sim |>
  filter(!is.na(Cc)) |>
  mutate(time_ss = time - win_start) |>
  filter(time_ss >= 0, time_ss <= 24) |>
  select(id, treatment, time_ss, Cc)
stopifnot(any(conc_df$time_ss == 0))   # a time-zero record is mandatory

dose_df <- cohort_ev |>
  filter(evid == 1L) |>
  mutate(time_ss = time - win_start) |>
  filter(time_ss >= 0, time_ss < 24) |>
  group_by(id, treatment) |>
  summarise(dose_total = sum(amt), time_ss = min(time_ss), .groups = "drop")

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time_ss | treatment + id,
                             concu = "ng/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, dose_total ~ time_ss | treatment + id,
                             doseu = "mg")

intervals <- data.frame(start = 0, end = 24,
                        cmax = TRUE, tmax = TRUE, cmin = TRUE,
                        auclast = TRUE, cav = TRUE, ctrough = TRUE)

# Match the paper's own NCA method (Methods 2.5, "linear-log trapezoidal rule").
# PKNCA.options() returns the previous values, so the old settings are restored
# after this chunk rather than leaking into later chunks.
old_opts <- PKNCA::PKNCA.options(auc.method = "lin up/log down")
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
stopifnot(nrow(as.data.frame(nca_res)) > 0L)

nca_res |>
  summary() |>
  knitr::kable(caption =
    "Steady-state NCA of the simulated cohort over the 24 h dosing day.")
```

## Comparison against the published NCA

Table 2 of the paper reports observed geometric-mean AUC24 and Ctrough by
formulation and genotype. Those were measured at the patients' own individual
clinical doses (median 5 and 3 mg/day IR-Tac in expressers and non-expressers),
not at the fixed 3 mg/day used above, so the comparison below re-simulates each
arm at the paper's own median dose for that group.

```{r nca-comparison}
tab2 <- tibble::tribble(
  ~treatment,                       ~dose_daily, ~form_lcp, ~expr, ~auclast, ~cmin,
  "IR-Tac, expresser",              5.00,  0L, 1L, 195, 4.9,
  "IR-Tac, non-expresser",          3.00,  0L, 0L, 212, 5.7,
  "LCP-Tac, expresser",             3.75,  1L, 1L, 232, 5.6,
  "LCP-Tac, non-expresser",         2.00,  1L, 0L, 199, 5.7
)

tab2_ev <- bind_rows(lapply(seq_len(nrow(tab2)), function(i) {
  r <- tab2[i, ]
  make_arm(r$dose_daily, form_lcp = r$form_lcp, expr = r$expr,
           n_sub = N_PER_ARM, id_offset = (i - 1L) * N_PER_ARM, obs_by = 0.25,
           label = r$treatment)
}))

sim2 <- rxode2::rxSolve(
  mod, tab2_ev,
  keep = c("treatment", "win_start"),
  returnType = "data.frame", atol = 1e-8, rtol = 1e-8, maxsteps = 1e6
)
if (is.null(sim2$id)) sim2$id <- 1L

conc2 <- sim2 |>
  filter(!is.na(Cc)) |>
  mutate(time_ss = time - win_start) |>
  filter(time_ss >= 0, time_ss <= 24) |>
  select(id, treatment, time_ss, Cc)
dose2 <- tab2_ev |>
  filter(evid == 1L) |>
  mutate(time_ss = time - win_start) |>
  filter(time_ss >= 0, time_ss < 24) |>
  group_by(id, treatment) |>
  summarise(dose_total = sum(amt), time_ss = min(time_ss), .groups = "drop")

nca2 <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc2, Cc ~ time_ss | treatment + id,
                   concu = "ng/mL", timeu = "h"),
  PKNCA::PKNCAdose(dose2, dose_total ~ time_ss | treatment + id, doseu = "mg"),
  intervals = data.frame(start = 0, end = 24, auclast = TRUE, cmin = TRUE)
))

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca2,
  reference = tab2 |> select(treatment, auclast, cmin),
  by        = "treatment",
  units     = c(auclast = "ng*h/mL", cmin = "ng/mL"),
  tolerance_pct = 20
)

knitr::kable(cmp, caption = paste(
  "Simulated vs observed NCA (Mohammed Ali 2025 Table 2 geometric means).",
  "* differs from the reference by more than 20%."),
  align = c("l", "l", "r", "r", "r"))
attr(cmp, "footnote")
```

The simulated exposures track the observed geometric means across all four
formulation-by-genotype arms. Residual differences are expected and are not
tuned away: Table 2 reports geometric means over patients receiving a *range*
of doses (the median is used here as a single representative dose), the
cohort's simulated Ctrough is sensitive to the large `Vc/F` variability
(53% IIV plus 59% IOV), and the observed values carry the 13.3% residual error
that the simulation reproduces only in expectation.

# Assumptions and deviations

* **Clock time of the simulated doses is unstated and is the largest residual
  uncertainty.** The circadian terms make every prediction depend on the clock
  time at which dosing starts, but the paper never reports it. This vignette
  uses `dose_hour = 0`, the literal reading of the paper's own definition of
  `TIME` ("minutes starting at midnight of the first PK profile") with the
  first dose at the `TIME` origin. This reproduces the direction and most of
  the magnitude of the day/night asymmetry (ratio 1.13 modelled vs 1.17 in
  Table S1) and leaves IR-Tac AUC24 about 3% above Table S1. A first-dose time
  near 01:30 reproduces both the IR-Tac total AUC (within 0.1%) and the
  day/night ratio (within 1%) simultaneously, but no value in the paper
  supports preferring it, so it was not adopted. LCP-Tac predictions are
  insensitive to this choice, because a full-cycle integral of the rhythm does
  not depend on where the cycle starts.
* **The paper's simulations are slightly short of steady state, which also
  contributes to the IR-Tac offset.** Methods 2.8 simulates "at least 10 days
  of treatment", but the terminal half-life implied by Table 3 (with
  `Vp/F` = 500 L) is 45.8 h, so 10 days is only 5.2 terminal half-lives and
  reaches roughly 97% of the steady-state AUC. This vignette simulates 30 days
  (15.7 half-lives, within 0.002% of steady state), so a small positive offset
  against Table S1 is expected on structural grounds and was not tuned away.
* **Scope of the `CL/F` circadian rhythm was inferred, not stated.** The paper
  says the `ka` rhythm belongs to IR-Tac but is silent on whether the `CL/F`
  rhythm applies to LCP-Tac records. It is applied to both, on the grounds that
  `CL/F` is one shared parameter of a joint model and diurnal metabolic
  capacity is a patient property. Gate 3 above shows this reading matches
  Table S1 roughly five times better than the alternative.
* **Placement of the random effect relative to the cosinor term was inferred.**
  Equation 3 reads `P = theta_1 + theta_AMP * cos(...)` and calls `theta_1` the
  "individual value". Adding an *absolute* amplitude to the individual value is
  not viable here: the IR-Tac `ka` carries 150.66 %CV IIV, so `ka_i` would fall
  below the 1.55 h^-1 amplitude in roughly 40% of subjects and the absorption
  rate would go negative for part of every day, which cannot be what produced
  the paper's 1000-subject simulations. The model therefore uses the standard
  NONMEM idiom `P_i(t) = (theta_mesor + theta_AMP * cos(...)) * exp(eta)`, in
  which the amplitude-to-mesor ratio is constant across subjects and both rates
  stay positive. For the typical subject the two readings are identical, so no
  validation target distinguishes them.
* **Equation 3 is printed with unbalanced parentheses.** As typeset,
  `P = theta_1 + theta_AMP*COS(2*pi/1440)*(TIME - theta_ACROPHASE))` would make
  `cos(2*pi/1440)` a constant near 1 and the parameter a linear ramp in time,
  which contradicts the surrounding text ("a cosine function with a period of
  24 h", "acrophase (time of the peak of the cosine function)") and the
  reported acrophases. The standard cosinor reading
  `theta_1 + theta_AMP * cos(2*pi*(TIME - ACROPHASE)/24)` is used.
* **Acrophase units.** Equation 3 puts `TIME` in minutes with a 1440-minute
  period, but Table 3 labels both acrophases in hours (17 h, 3.13 h). The
  implementation works in hours with a 24 h period, which is the same function.
* **`Vp/F` is fixed at 500 L**, as in the paper (Table 3 "500 FIX"); no
  uncertainty is available for it.
* **Number of occasions for the IOV terms is inferred from the design.** The
  paper reports single IOV magnitudes for `CL/F` and `Vc/F` without stating how
  many occasions were used. The design fixes it at two, one rich 24 h profile
  per formulation, so `OCC = 1` is the IR-Tac profile and `OCC = 2` the
  LCP-Tac profile, with occasion 2's variance fixed equal to occasion 1
  (NONMEM `$OMEGA BLOCK(1) SAME`). Because `OCC` and `FORM_TAC_ENVARSUS` are
  one-to-one in this crossover, the IOV and the formulation effect on `F` are
  partially aliased; a user applying this model to a dataset with more
  occasions per formulation should revisit the mapping.
* **Residual error is coded as purely proportional.** Table 3 labels the
  13.30% term "Combined residual error", but no additive component is reported
  anywhere in the paper and Results 3.2.1 states the proportional model was
  selected. A single proportional term is used.
* **`Lag-Time LCP-Tac` is 1.4 h**, per Table 3 and its bootstrap CI
  (1.32-1.57); the Discussion quotes 1.42 h for the same parameter.
* **Covariates screened but not retained.** Body weight, BMI, age, sex and
  hematocrit were all tested and none reached significance (Results 3.2.2), so
  they are recorded in `covariatesDataExcluded` rather than `covariateData` and
  the model has no size scaling. Users applying it outside the narrow
  demographic range of this 30-patient cohort should not assume weight
  independence generalises.
* **The single `CYP3A5*1/*1` patient was pooled with heterozygotes** by the
  authors (Results 3.1), so `CYP3A5_EXPR = 1` covers both `*1/*3` and `*1/*1`
  and the model cannot distinguish homozygous from heterozygous expressers.
* **`checkModelConventions()` reports no issues** beyond an informational note
  that `units$dosing` (mg) and `units$concentration` (ng/mL) differ in
  magnitude; the required factor of 1000 is applied explicitly in
  `Cc <- 1000 * central / vc`.
* **Original observed data are not publicly available**, so all figures use
  virtual cohorts and typical-value profiles rather than the trial's own
  concentrations. The quantitative comparisons target the paper's published
  tables (Table 2 observed NCA, Table 3 parameter estimates, Table S1
  simulations) and the four derived `CL/F` values in the Discussion.
