---
title: "Copanlisib (Morcos 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Copanlisib (Morcos 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Morcos_2023_copanlisib"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1002/psp4.13000>
* Supplement (Tables S1-S3, Figures S1-S4): <https://doi.org/10.1002/psp4.13000> (Supporting Information, file `PSP4-12-1666-s001.docx`)

Copanlisib is an intravenous pan-class-I PI3K inhibitor approved for relapsed
follicular lymphoma. Morcos 2023 pooled 5958 plasma concentrations from 712
patients across nine phase I-III studies into a single population PK model, and
then used that model's individual exposure predictions to drive
exposure-response analyses in the phase III CHRONOS-3 trial.

The packaged model is the paper's **final population PK covariate model**
(Table 2). The exposure-response layer is *not* packaged; see
[Assumptions and deviations](#assumptions-and-deviations) for why.

# Population

```{r population, echo = FALSE}
pop <- ui$population
tibble::tibble(
  Field = names(pop),
  Value = vapply(pop, function(x) paste(as.character(x), collapse = "; "), character(1))
) |>
  knitr::kable(caption = "Population metadata carried by the model file (Morcos 2023 Table 1, pooled PopPK column, and Table S1).")
```

The pooled analysis population is 712 adults with advanced solid tumors or
non-Hodgkin lymphoma, median age 63 years (range 20-91), median body weight
70.0 kg (range 41.1-165), 52.4% female (Morcos 2023 Table 1). Copanlisib was
given as a 1-h intravenous infusion on days 1, 8 and 15 of a 28-day cycle, at
0.1-1.2 mg/kg in the dose-escalation studies and at a 60 mg flat dose in the
phase II and phase III studies. Concentrations were measured by validated
LC/MS with an LLOQ of 2 ng/mL; 276 of 5958 observations were below that limit
and were handled with the Beal M3 method.

The simulations below reproduce the **CHRONOS-3** subpopulation
(Bayer study 17067, NCT02367040), because that is the cohort for which the
paper publishes exposure percentiles. Its baseline distribution comes from the
`Study 17067` column of Table 1 (n = 447): 47.7% female, 15.7% with any hepatic
impairment (15.2% mild + 0.4% moderate), and 7.8% enrolled at Japanese sites.

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Morcos_2023_copanlisib.R` carries an in-file
comment naming its source. They are collected here for review.

| Equation / parameter | Value | Source location |
|---|---|---|
| Three-compartment IV disposition, first-order elimination from central | n/a | Results, "PopPK meta-analyses"; schematic Figure S1 |
| Categorical covariate form `P = Ptv * (1 + N1*th1 + N2*th2 + ...) * exp(eta)` | n/a | Methods, "Copanlisib PopPK modeling" (third displayed equation) |
| `lcl` (CL) | 22.2 L/h | Table 2, `CL pop` (RSE 3.18%, 95% CI 20.8-23.5) |
| `lvc` (V1) | 92.1 L | Table 2, `V1 pop` (RSE 7.47%, 95% CI 78.6-106) |
| `lq` (Q2) | 79.3 L/h | Table 2, `Q2` (RSE 1.58%) |
| `lvp` (V2) | 508 L | Table 2, `V2` (RSE 2.54%) |
| `lq2` (Q3) | 7.34 L/h | Table 2, `Q3` (RSE 6.96%) |
| `lvp2` (V3) | 522 L | Table 2, `V3` (RSE 4.26%) |
| `e_conmed_rifampicin_cl` | 1.91 | Table 2 `Theta RIFCL`; Results "increased CL by 191%" |
| `e_conmed_itraconazole_cl` | -0.361 | Table 2 `Theta ITRACL`; Results "decreased CL by 36.1%" |
| `e_study_chronos3_cl` | -0.184 | Table 2 `Theta 17067CL`; Results "18.4% lower CL" |
| `e_sexf_cl` | -0.167 | Table 2 `Theta SEXCL`; Results "females had 16.7% lower CL" |
| `e_hepimp_cl` | -0.192 | Table 2 `Theta NCICL`; Results "19.2% lower CL" |
| `e_region_japan_cl` | -0.204 | Table 2 `Theta JAPCL`; Results "Japan had 20.4% lower CL" |
| `e_sexf_vc` | -0.429 | Table 2 `Theta SEXV1`; Results "females had 42.9% lower V1" |
| `e_conmed_rifampicin_vc` | 1.08 | Table 2 `Theta RIFV1`; Results "rifampin increased V1 by 108%" |
| `etalcl` variance | 0.124 | Table 2 OMEGA on `CL pop` (CV 36.3%, shrinkage 22.5%) |
| `etalvc` variance | 0.846 | Table 2 OMEGA on `V1 pop` (CV 115%, shrinkage 27.3%) |
| `expSd_early` | sqrt(5.10) = 2.2583 | Table 2 SIGMA, "first 20 min of an infusion" (CV 1280%) |
| `expSd_phase12` | sqrt(0.176) = 0.4195 | Table 2 SIGMA, "phase I and phase II ... after first 20 min" (CV 43.9%) |
| `expSd_phase3` | sqrt(0.632) = 0.7950 | Table 2 SIGMA, "phase III for study 17067 ... after first 20 min" (CV 93.8%) |
| Log-additive residual = `lnorm()` | n/a | Table 2 footnote g |
| Reference patient (all indicators 0) | n/a | Table 2 footnotes d and e |

The paper's reference patient is a **male, in a non-Japanese phase I or phase II
study, without rifampicin or itraconazole comedication, with normal hepatic
function**. Setting every covariate indicator to 0 in the packaged model
recovers that patient exactly.

# Variance components reproduce Table 2

Table 2 reports each variance twice: as `omega^2` / `sigma^2`, and as a
percent CV derived by `100 * sqrt(exp(v) - 1)` (footnotes f and g). The
packaged model stores standard deviations, so back-transforming them must
regenerate the published CV column. This is a deterministic identity - no
simulation, no cohort - so it is asserted tightly.

```{r variance-identities}
iniDf <- ui$iniDf

get_est <- function(nm) {
  v <- iniDf$est[iniDf$name == nm]
  stopifnot(length(v) == 1L)
  v
}

cv_pct <- function(variance) 100 * sqrt(exp(variance) - 1)

variance_check <- tibble::tibble(
  Component = c("IIV on CL", "IIV on V1",
                "Residual, first 20 min of infusion",
                "Residual, phase I/II beyond 20 min",
                "Residual, phase III beyond 20 min"),
  `Table 2 variance` = c(0.124, 0.846, 5.10, 0.176, 0.632),
  `Model value` = c(
    get_est("etalcl"),           # stored as a variance
    get_est("etalvc"),           # stored as a variance
    get_est("expSd_early")^2,    # stored as a log-scale SD
    get_est("expSd_phase12")^2,
    get_est("expSd_phase3")^2
  ),
  `Table 2 CV (%)` = c(36.3, 115, 1280, 43.9, 93.8)
) |>
  mutate(
    `Model CV (%)` = cv_pct(`Model value`),
    `CV rel. diff (%)` = 100 * (`Model CV (%)` - `Table 2 CV (%)`) / `Table 2 CV (%)`
  )

knitr::kable(variance_check, digits = c(0, 4, 4, 1, 1, 2),
             caption = "Back-transformed variance components against the published CV column of Morcos 2023 Table 2.")

# The published CV column is rounded to 3 significant figures, so 0.5% is the
# tightest bound that rounding alone permits; anything larger means a
# mis-transcribed variance or a variance/SD confusion.
stopifnot(max(abs(variance_check$`CV rel. diff (%)`)) < 0.5)
```

# Covariate effects reproduce Table 2

Each of the eight retained covariates is a binary indicator entering as a
multiplicative `(1 + theta)` factor. Solving the typical-value model
(`zeroRe()`) once per single-covariate configuration and dividing each
clearance by the reference patient's clearance must return `1 + theta`
exactly.

```{r covariate-identities}
mod <- readModelDb("Morcos_2023_copanlisib")
mod_typ <- rxode2::zeroRe(mod)

cov_names <- c("SEXF", "HEPIMP", "REGION_JAPAN", "CONMED_RIFAMPICIN",
               "CONMED_ITRACONAZOLE", "STUDY_CHRONOS3")

# One subject per configuration: subject 1 is the reference patient (all
# indicators 0), then one subject per covariate turned on alone.
probe_cov <- matrix(0, nrow = 1 + length(cov_names), ncol = length(cov_names),
                    dimnames = list(NULL, cov_names))
for (i in seq_along(cov_names)) probe_cov[i + 1L, cov_names[i]] <- 1

probe_subj <- tibble::as_tibble(probe_cov) |>
  mutate(
    id = seq_len(n()),
    config = c("Reference (all indicators 0)",
               "Female", "Any hepatic impairment", "Japan",
               "Rifampicin", "Itraconazole", "CHRONOS-3")
  )

# 60 mg over a 1-h infusion into `central`; observations on the ODE state.
probe_dose <- probe_subj |>
  mutate(time = 0, amt = 60, evid = 1L, cmt = "central", rate = 60)
probe_obs <- probe_subj |>
  tidyr::crossing(time = c(0.5, 1, 2)) |>
  mutate(amt = NA_real_, evid = 0L, cmt = "central", rate = NA_real_)

probe_ev <- bind_rows(probe_dose, probe_obs) |>
  select(id, time, amt, evid, cmt, rate, all_of(cov_names), config) |>
  arrange(id, time, desc(evid))

probe_sim <- rxode2::rxSolve(mod_typ, events = probe_ev, keep = "config") |>
  as.data.frame()

probe_par <- probe_sim |>
  group_by(id, config) |>
  summarise(cl = unique(round(cl, 10)), vc = unique(round(vc, 10)), .groups = "drop")

ref_cl <- probe_par$cl[probe_par$config == "Reference (all indicators 0)"]
ref_vc <- probe_par$vc[probe_par$config == "Reference (all indicators 0)"]

cov_check <- probe_par |>
  filter(config != "Reference (all indicators 0)") |>
  mutate(
    `CL factor` = cl / ref_cl,
    `V1 factor` = vc / ref_vc
  ) |>
  select(Configuration = config, `CL factor`, `V1 factor`) |>
  mutate(
    `Expected CL factor` = 1 + c(-0.167, -0.192, -0.204, 1.91, -0.361, -0.184),
    `Expected V1 factor` = 1 + c(-0.429, 0, 0, 1.08, 0, 0)
  )

knitr::kable(cov_check, digits = 4,
             caption = "Typical-value covariate factors against 1 + theta from Morcos 2023 Table 2.")

# Deterministic algebra; the only slack is double-precision round-off.
stopifnot(
  max(abs(cov_check$`CL factor` - cov_check$`Expected CL factor`)) < 1e-8,
  max(abs(cov_check$`V1 factor` - cov_check$`Expected V1 factor`)) < 1e-8
)

# The reference patient's typical values must be Table 2's CL pop and V1 pop.
stopifnot(
  abs(ref_cl - 22.2) < 1e-8,
  abs(ref_vc - 92.1) < 1e-8
)
```

# Disposition of a single 60 mg dose

Before turning to the intermittent regimen, the typical-value profile after one
60 mg infusion establishes the terminal half-life and confirms the fundamental
identity `AUC(0-inf) = dose / CL`.

```{r single-dose}
sd_times <- sort(unique(c(
  seq(0, 4, by = 0.05), seq(4, 24, by = 0.25),
  seq(24, 168, by = 2), seq(168, 1008, by = 8)
)))

sd_ev <- bind_rows(
  tibble::tibble(id = 1L, time = 0, amt = 60, evid = 1L, cmt = "central", rate = 60),
  tibble::tibble(id = 1L, time = sd_times, amt = NA_real_, evid = 0L,
                 cmt = "central", rate = NA_real_)
) |>
  mutate(SEXF = 0, HEPIMP = 0, REGION_JAPAN = 0, CONMED_RIFAMPICIN = 0,
         CONMED_ITRACONAZOLE = 0, STUDY_CHRONOS3 = 0) |>
  arrange(time, desc(evid))

sd_sim <- rxode2::rxSolve(mod_typ, events = sd_ev) |> as.data.frame()

# rxSolve DROPS the id column entirely when the event table holds a single
# subject, and PKNCA's grouping formula below needs it. Restore it rather than
# assuming it is there.
if (is.null(sd_sim$id)) sd_sim$id <- 1L

# Concentrations must stay non-negative; a negative far tail would make
# PKNCA's log-down trapezoid produce NaN (see pknca-recipes.md).
stopifnot(all(sd_sim$Cc >= 0))

sd_sim |>
  filter(time > 0) |>
  ggplot(aes(time, Cc)) +
  geom_line(linewidth = 0.7) +
  scale_x_continuous(breaks = seq(0, 1008, by = 168)) +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Copanlisib concentration (mg/L)",
    title = "Typical-value profile, single 60 mg 1-h intravenous infusion",
    caption = "Three-compartment disposition of Morcos 2023 Table 2, reference patient."
  ) +
  theme_bw()
```

```{r single-dose-nca}
sd_conc <- sd_sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc) |>
  mutate(treatment = "60 mg single dose")

sd_dose <- tibble::tibble(id = 1L, time = 0, amt = 60,
                          treatment = "60 mg single dose")

sd_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(as.data.frame(sd_conc), Cc ~ time | treatment + id,
                   concu = "mg/L", timeu = "h"),
  PKNCA::PKNCAdose(as.data.frame(sd_dose), amt ~ time | treatment + id,
                   doseu = "mg"),
  intervals = data.frame(start = 0, end = Inf,
                         cmax = TRUE, tmax = TRUE,
                         auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE)
))

sd_tbl <- as.data.frame(sd_res$result) |>
  select(PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

knitr::kable(
  sd_tbl |>
    dplyr::select(cmax, tmax, auclast, aucinf.obs, half.life) |>
    dplyr::rename(
      "Cmax (mg/L)"         = cmax,
      "Tmax (h)"            = tmax,
      "AUClast (mg*h/L)"    = auclast,
      "AUC0-inf (mg*h/L)"   = aucinf.obs,
      "Terminal t1/2 (h)"   = half.life
    ),
  digits = 3,
  caption = "PKNCA on the typical-value single-dose profile."
)

auc_inf <- sd_tbl$aucinf.obs
thalf   <- sd_tbl$half.life

# Fail loudly rather than silently reporting NA downstream: an NA half-life
# would mean PKNCA found too few post-peak points to fit lambda-z, which is a
# problem with the sampling grid above, not a property of the model.
stopifnot(is.finite(auc_inf), is.finite(thalf))

# Internal identity: for a linear model with no IIV and no residual error,
# AUC(0-inf) is exactly dose / CL. The only error is trapezoidal.
stopifnot(abs(auc_inf - 60 / 22.2) / (60 / 22.2) < 0.01)

# The 1008 h window is many terminal half-lives long, so AUClast has
# effectively converged onto AUC(0-inf); assert >= because auclast can only
# undershoot aucinf.obs.
stopifnot(sd_tbl$auclast <= auc_inf,
          (auc_inf - sd_tbl$auclast) / auc_inf < 0.01)
```

The model's terminal half-life is `r round(thalf, 1)` h. Morcos 2023's
Introduction quotes "a half-life of around 38 h" at the maximum tolerated dose,
but that figure comes from the earlier phase I analysis (reference 5 of the
paper) and is a *reported* rather than a *modelled* value; the present
three-compartment model, which resolves a deep peripheral compartment with
Q3 = 7.34 L/h and V3 = 522 L, necessarily supports a longer terminal slope.
The paper itself reports no half-life for this model, so this is a
cross-analysis observation, not a discrepancy - it is recorded in
[Assumptions and deviations](#assumptions-and-deviations) and is deliberately
excluded from the assertion gate.

# Virtual CHRONOS-3 cohort

```{r cohort}
# set.seed() seeds R's RNG, which draws the covariate configuration below. It
# does NOT seed rxode2's simulation RNG, and rxode2 partitions its streams per
# solver thread -- so the etas drawn downstream differ between a 2-core CI
# runner and a many-thread workstation. Every assertion below is therefore
# written to hold for any cohort this model can produce.
set.seed(20231101)

n_sub <- 200L   # cap is 200 participants per arm

# Baseline proportions from Morcos 2023 Table 1, column "Study 17067" (n = 447,
# the CHRONOS-3 exposure-response population): 213/447 = 47.7% female;
# (68 + 2)/447 = 15.7% with any hepatic impairment; 35/447 = 7.8% Japan.
#
# Subgroup membership is assigned with EXACT counts rather than rbinom(), so
# every subgroup has a fixed, reproducible size on any machine. With only ~8%
# Japan prevalence, binomial assignment would let the Japanese subgroup range
# from about 8 to 25 subjects between runs, and the subgroup geometric mean
# asserted below would inherit that extra noise for no benefit.
assign_exact <- function(n, k) {
  v <- integer(n)
  v[sample.int(n, k)] <- 1L
  v
}

subj <- tibble::tibble(
  id                  = seq_len(n_sub),
  SEXF                = assign_exact(n_sub, 95L),   # 47.7% of 200 = 95.4
  HEPIMP              = assign_exact(n_sub, 31L),   # 15.7% of 200 = 31.4
  REGION_JAPAN        = assign_exact(n_sub, 16L),   #  7.8% of 200 = 15.6
  CONMED_RIFAMPICIN   = 0,   # given only in the dedicated DDI study 16270
  CONMED_ITRACONAZOLE = 0,   # given only in the dedicated DDI study 16270
  STUDY_CHRONOS3      = 1,
  treatment           = "Copanlisib 60 mg, days 1/8/15"
)

stopifnot(sum(subj$SEXF) == 95L, sum(subj$HEPIMP) == 31L,
          sum(subj$REGION_JAPAN) == 16L)

# AUC(0-168)nd is defined in Methods, "Determination of copanlisib exposure
# metrics": the AUC from 0 to 168 h AFTER THE THIRD 60 mg nominal dose in a
# sequence of three doses of 60 mg each one week apart. So dose at 0, 168 and
# 336 h and integrate over [336, 504].
dose_times <- c(0, 168, 336)
auc_start  <- 336
auc_end    <- 504

# Dense sampling around each infusion plus a regular backbone, so the
# trapezoidal AUC over [336, 504] resolves the distribution phase.
fine <- c(0, 0.25, 0.5, 0.75, 1, 1.1, 1.25, 1.5, 2, 2.5, 3, 4, 5, 6, 8, 11, 14, 18, 24)
obs_times <- sort(unique(c(
  as.vector(outer(dose_times, fine, "+")),
  seq(0, auc_end, by = 4),
  auc_start, auc_end
)))
obs_times <- obs_times[obs_times <= auc_end]

events <- bind_rows(
  subj |>
    tidyr::crossing(time = dose_times) |>
    mutate(amt = 60, evid = 1L, cmt = "central", rate = 60),
  subj |>
    tidyr::crossing(time = obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central", rate = NA_real_)
) |>
  select(id, time, amt, evid, cmt, rate,
         SEXF, HEPIMP, REGION_JAPAN, CONMED_RIFAMPICIN, CONMED_ITRACONAZOLE,
         STUDY_CHRONOS3, treatment) |>
  arrange(id, time, desc(evid))

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod, events = events,
  keep = c("SEXF", "HEPIMP", "REGION_JAPAN", "treatment")
) |>
  as.data.frame()

# rxSolve returns Cc (== ipredSim, NO residual error) alongside `sim` (which
# DOES carry the residual). Everything below deliberately uses Cc: the
# published exposure metrics are model-predicted AUCs, not assayed values.
stopifnot(isTRUE(all.equal(sim$Cc, sim$ipredSim)))
stopifnot(all(sim$Cc >= 0), !anyNA(sim$Cc))
```

```{r figure-1}
# Companion to Figure 1b of Morcos 2023 (prediction-corrected VPC of copanlisib
# PK in CHRONOS-3). Shown here over the third dosing interval, where the
# published AUC(0-168)nd is defined.
sim |>
  filter(time >= auc_start, time <= auc_end) |>
  mutate(tad = time - auc_start) |>
  group_by(tad) |>
  summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  filter(tad > 0) |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.7) +
  scale_x_continuous(breaks = seq(0, 168, by = 24)) +
  scale_y_log10() +
  labs(
    x = "Time after third dose (h)", y = "Copanlisib concentration (mg/L)",
    title = "Simulated CHRONOS-3 exposure over the third dosing interval",
    caption = paste("Median with 5th-95th percentile band, n =", n_sub,
                    "virtual patients. Companion to Figure 1b of Morcos 2023.")
  ) +
  theme_bw()
```

# PKNCA validation

```{r pknca}
# Only !is.na(Cc) -- adding time > 0 or Cc > 0 would drop the anchor row that
# PKNCA needs at the start of the interval.
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)

# Guarantee a time-zero row per subject. This is an intravenous infusion, so
# the pre-dose concentration is 0.
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(
  as.data.frame(sim_nca), Cc ~ time | treatment + id,
  concu = "mg/L", timeu = "h"
)

dose_df <- events |>
  filter(evid == 1L) |>
  select(id, time, amt, treatment)

dose_obj <- PKNCA::PKNCAdose(
  as.data.frame(dose_df), amt ~ time | treatment + id, doseu = "mg"
)

# AUC(0-168)nd = AUC over the interval that follows the third 60 mg dose.
intervals <- data.frame(
  start   = auc_start,
  end     = auc_end,
  cmax    = TRUE,
  tmax    = TRUE,
  auclast = TRUE,
  cav     = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_tbl <- as.data.frame(nca_res$result) |>
  select(id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

# Convert the model's mg/L*h onto the paper's ug*h/L (1 mg/L = 1000 ug/L).
nca_tbl <- nca_tbl |> mutate(auc_ug_h_L = auclast * 1000)

auc_q <- quantile(nca_tbl$auc_ug_h_L, c(0.05, 0.50, 0.95))
```

## Comparison against published exposure

Morcos 2023 reports AUC(0-168)nd for copanlisib-treated CHRONOS-3 patients in
two places that agree with each other: Figure 3b prints a median of 3720
ug*h/L with 5th-95th percentiles of 2650-5770, and the Discussion gives "the
fifth and 95th percentiles of AUC[0-168]nd in CHRONOS-3 were 2647 and
5766 ng*h/mL" (1 ng/mL = 1 ug/L, so the units are the same).

```{r nca_compare}
# AUC(0-168)nd is the ONLY NCA quantity this paper publishes for the model.
# Cmax and Tmax are never reported, and the paper's Cavg,2wk / Cavg,4wk /
# Cavg,8wk medians (13.9 / 13.8 / 13.1 ug/L, Figure 3b) are moving averages
# over each patient's ACTUAL dosing history -- which includes the week off
# every cycle and the dose interruptions that 75.2% of CHRONOS-3 patients
# experienced -- so they are not comparable to a nominal three-dose schedule
# and are deliberately excluded from this table.
published <- tibble::tibble(
  treatment = "Copanlisib 60 mg, days 1/8/15",
  auclast   = 3.720      # 3720 ug*h/L (Figure 3b) expressed in mg*h/L
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "treatment",
  params        = "auclast",
  units         = c(auclast = "mg*h/L"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated AUC(0-168)nd (median of 200 virtual CHRONOS-3 patients) vs. Morcos 2023 Figure 3b. * differs from reference by >20%."
)
```

```{r exposure-distribution}
exposure_check <- tibble::tibble(
  Statistic   = c("5th percentile", "Median", "95th percentile"),
  `Published (ug*h/L)` = c(2650, 3720, 5770),
  `Simulated (ug*h/L)` = as.numeric(auc_q)
) |>
  mutate(`Rel. diff (%)` = 100 * (`Simulated (ug*h/L)` - `Published (ug*h/L)`) /
           `Published (ug*h/L)`)

knitr::kable(exposure_check, digits = c(0, 0, 0, 1),
             caption = "AUC(0-168)nd distribution against Morcos 2023 Figure 3b.")

med_pct <- exposure_check$`Rel. diff (%)`[exposure_check$Statistic == "Median"]
```

The **median** is the structural check, and it is the one asserted tightly: a
mis-transcribed clearance, dose, covariate coefficient or unit moves the whole
distribution by tens of percent.

```{r assert-median}
# AUC(0-168)nd = dose / CL to better than 1% (by superposition the interval
# spans the whole of a single dose's disposition), so the median AUC is pinned
# by the median individual clearance. Subgroup sizes are fixed by construction,
# so the only noise left is the eta draw: the sample median of 200 log-normal
# draws with sigma = sqrt(0.124) = 0.352 has a log-scale standard error of
# 1.2533 * 0.352 / sqrt(200) = 0.031, i.e. +/- 9.4% at three standard errors.
# 15% therefore sits outside the achievable noise while still going red on any
# structural error worth catching: dropping the -18.4% CHRONOS-3 clearance
# effect alone moves the median by 22.5%, and a unit or dose error moves it by
# orders of magnitude.
stopifnot(abs(med_pct) < 15)
```

The **tails are deliberately not gated**, and the table above shows why: the
simulated 5th-95th spread is wider than the published one. That is expected and
is not a transcription problem. The published percentiles are computed from
*individual* (empirical Bayes) parameter estimates, which Table 2 reports as
22.5% shrunk on CL and 27.3% shrunk on V1; shrinkage pulls individual estimates
toward the typical value and narrows the observed spread. A forward simulation
draws etas from the full estimated OMEGA and therefore reproduces the
*estimated* population variability rather than the *shrunken* empirical one. For
9.5% of the exposure-response patients the paper did not even have PK
observations and used population parameters, narrowing the published spread
further.

```{r assert-spread}
sim_ratio <- unname(auc_q[3] / auc_q[1])
pub_ratio <- 5770 / 2650      # = 2.18

# Absolute bounds, not a race between two noisy statistics. Pure IIV on CL
# (omega^2 = 0.124) predicts a 5th-95th ratio of
# exp(2 * 1.645 * sqrt(0.124)) = 3.18, with the covariates widening it a
# little further. The 5th and 95th percentiles of n = 200 each carry a
# log-scale standard error of about 0.053, so the ratio's three-sigma range is
# roughly 2.5 to 4.0. Bounds of 2.4 and 5.0 sit outside that range at both
# ends and can still go red: halving or doubling the encoded IIV variance
# moves this ratio to 1.8 or 5.7 respectively.
stopifnot(sim_ratio > 2.4, sim_ratio < 5)

# And the qualitative claim of the paragraph above -- forward simulation
# cannot shrink, so the simulated spread exceeds the shrunken published one.
# The margin here is structural (3.18 vs 2.18, a 46% gap), not a coin flip.
stopifnot(sim_ratio > pub_ratio)
```

# Covariate influence on exposure (Figure 2)

Figure 2 of Morcos 2023 is a forest plot of geometric-mean AUC(0-168)nd ratios
across CHRONOS-3 subgroups. The paper makes two quantitative claims about it:

1. Japan versus Europe is the only comparison outside the 0.8-1.25
   bioequivalence range, with a geometric mean ratio of 1.34 (90% CI 1.27,
   1.43).
2. "No covariate showed exposure differences greater than around 35%."

The model's own contribution to the Japan effect is deterministic and can be
checked exactly: `1 / (1 - 0.204) = 1.2563`.

```{r forest}
nca_cov <- nca_tbl |>
  left_join(distinct(subj, id, SEXF, HEPIMP, REGION_JAPAN), by = "id")

gm <- function(x) exp(mean(log(x)))

gmr <- function(flag) {
  a <- nca_cov$auc_ug_h_L[flag == 1]
  b <- nca_cov$auc_ug_h_L[flag == 0]
  if (length(a) < 5 || length(b) < 5) return(NA_real_)
  gm(a) / gm(b)
}

forest <- tibble::tibble(
  Comparison = c("Female : Male",
                 "Any hepatic impairment : Normal",
                 "Japan : Non-Japan"),
  `Geometric mean ratio` = c(gmr(nca_cov$SEXF),
                             gmr(nca_cov$HEPIMP),
                             gmr(nca_cov$REGION_JAPAN)),
  `Model factor` = c(1 / (1 - 0.167), 1 / (1 - 0.192), 1 / (1 - 0.204))
)

knitr::kable(forest, digits = 3,
             caption = "Simulated subgroup geometric-mean AUC(0-168)nd ratios against the deterministic model factor 1/(1 + theta).")

japan_gmr <- forest$`Geometric mean ratio`[forest$Comparison == "Japan : Non-Japan"]
```

```{r assert-forest}
# The deterministic model factor is exact algebra and is asserted as such.
stopifnot(abs(1 / (1 - 0.204) - 1.2563) < 5e-4)

# The simulated ratio is a geometric mean over a 16-subject subgroup, so it
# carries real eta sampling noise on top of the exact 1.256 model factor: the
# log-ratio standard error is sqrt(1/16 + 1/184) * 0.352 = 0.092, giving a
# three-sigma range of roughly 0.95 to 1.65, widened a little further by which
# of the 16 happen to also be female. Band the model factor, not the published
# 1.34 -- and note the band still goes red on a sign-flipped Japan
# coefficient, which would land near 0.80.
stopifnot(japan_gmr > 0.85, japan_gmr < 1.9)

# Claim 2 of the paper: no covariate moves exposure by more than ~35%. Every
# retained effect in the CHRONOS-3 setting (rifampicin and itraconazole are
# absent from that trial) is bounded by the Japan effect.
chronos3_factors <- c(1 / (1 - 0.167), 1 / (1 - 0.192), 1 / (1 - 0.204))
stopifnot(max(chronos3_factors) < 1.35)
```

The simulated Japan:non-Japan ratio for this cohort is
`r sprintf("%.3f", japan_gmr)`. It should be read against the model's *exact*
factor of 1.256, not against the paper's 1.34: with only 16 Japanese subjects
the simulated ratio scatters around 1.256 by roughly +/- 0.3, and it is
additionally displaced by however many of those 16 also drew the female
indicator, since sex carries a 1.20-fold effect on exposure of its own. A
single draw landing near 1.34 is therefore not evidence for or against the
published value.

The published 1.34 is a Japan-versus-*Europe* comparison of two observed
subgroups whose composition differs in the other retained covariates as well,
so it is not a pure covariate contrast and exceeds the model's own 1.256 term.
What *is* reproduced, and is deterministic, is the paper's structural
conclusion: 1.256 is the largest single covariate factor operating in
CHRONOS-3 (rifampicin and itraconazole never occur in that trial), and it is
the only one above the 1.25 bioequivalence bound - which is exactly why Figure
2 shows Japan as the sole subgroup outside 0.8-1.25.

# Assumptions and deviations {#assumptions-and-deviations}

- **The exposure-response layer is not packaged.** Morcos 2023 has a second
  half - multivariate Cox proportional hazards models for progression-free
  survival, time to serious adverse event and time to grade >= 3 treatment
  emergent adverse event, plus multivariate logistic regressions for objective
  response rate and for individual safety events. None of these is encodable as
  an rxode2 model, for a reason that is structural rather than a reporting gap:
  - The Cox models are **semiparametric**. Their baseline hazard `lambda0(t)`
    is a nonparametric step function estimated from the data, is never reported
    (and by construction cannot be reported as a small parameter set), so
    absolute event times cannot be simulated - only hazard *ratios* are
    identified.
  - The logistic models report **odds ratios only**. Figures 4b, 5b and 5d
    print each covariate's odds ratio or hazard ratio with its 95% CI, but no
    intercept `beta0`, so absolute event probabilities are not recoverable.

  The figure panels were inspected directly for these values before this
  conclusion was recorded (Figures 3b and 4b were rendered at 200 dpi from the
  publisher PDF, which carries them as vector text). They contain point
  estimates and confidence intervals for every covariate, and no baseline
  hazard or intercept. The reported effects are preserved in prose here for
  provenance: PFS hazard ratios were 0.128 (95% CI 0.0317, 0.519) for Japan,
  0.567 (0.421, 0.762) for above-median rituximab exposure before the fourth
  infusion, and 0.451 (0.336, 0.605) for copanlisib versus placebo; ORR odds
  ratios were 2.32 (1.53, 3.52) for follicular-lymphoma histology and 3.25
  (2.12, 4.98) for copanlisib versus placebo.

- **The M3 method for below-quantification-limit data is an estimation device
  only.** 276 of 5958 observations were below the 2 ng/mL LLOQ and were handled
  by the Beal M3 likelihood. Forward simulation has no counterpart, so the
  packaged model returns continuous concentrations at all times.

- **Body weight is documented but not used.** Allometric scaling on body weight
  was investigated (Methods) but did not survive the paper's backward
  elimination and does not appear in Table 2. It is recorded in the model file's
  `covariatesDataExcluded` list along with age, serum albumin, eGFR and the four
  non-Japan region indicators, so that the paper's covariate search is not lost,
  but the model applies no weight scaling. The Discussion attributes the lower
  clearance in Japanese patients to "general differences in body weight",
  meaning the `REGION_JAPAN` term partly stands in for a weight effect that was
  never estimated separately.

- **`STUDY_CHRONOS3` is confounded with rituximab comedication.** Rituximab was
  co-administered only in study 17067, so Table S2 records that a
  rituximab-comedication covariate is entirely confounded with the study
  indicator. The paper declines to attribute the 18.4% clearance reduction to
  rituximab, and this model follows it: the effect is labelled as a study
  effect, not a drug-interaction effect.

- **The terminal half-life is not a published value for this model.** The
  packaged three-compartment model gives a terminal half-life of
  `r round(thalf, 1)` h. The "around 38 h" quoted in the paper's Introduction
  comes from the earlier phase I analysis (its reference 5), which used a
  different structural model and a shorter sampling window. Morcos 2023 reports
  no half-life for its own model, so this is not asserted.

- **Cohort covariates are marginally correct but mutually independent.**
  `SEXF`, `HEPIMP` and `REGION_JAPAN` are assigned at exactly the Table 1
  marginal counts of the CHRONOS-3 column, but independently of one another,
  because the paper publishes only marginals and no joint distribution. Any
  real association between them - most relevantly a sex composition that
  differs by region - is therefore absent from the virtual cohort. This is why
  the simulated Japan:non-Japan geometric-mean ratio should be compared against
  the model's own 1.256 factor rather than against the paper's observed
  Japan:Europe ratio of 1.34, which carries those composition differences
  inside it.

- **Rifampicin and itraconazole are set to 0 throughout.** Both are time-varying
  covariates that were non-zero only within the dedicated drug-interaction study
  16270 (Table S2), which is not part of the CHRONOS-3 cohort simulated here.
  Their coefficients are still packaged and are exercised by the deterministic
  covariate-factor check above.
