---
title: "Acetaminophen and ibuprofen formulation and food effects (Morse 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Acetaminophen and ibuprofen formulation and food effects (Morse 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Morse 2022 fitted two independent population PK models to the same pooled
healthy-volunteer data set, one for acetaminophen and one for ibuprofen. Both
are packaged here and share this article.

* Citation: `r rxode2::rxode(readModelDb("Morse_2022_acetaminophen"))$reference`
* Acetaminophen model: `r rxode2::rxode(readModelDb("Morse_2022_acetaminophen"))$description`
* Ibuprofen model: `r rxode2::rxode(readModelDb("Morse_2022_ibuprofen"))$description`
* Article (open access): <https://doi.org/10.1007/s13318-022-00766-9>
* Electronic supplementary material (correlation matrices, Table S2; the
  pharmacodynamic parameters used for the pain-score simulations, Table S3):
  <https://static-content.springer.com/esm/art%3A10.1007%2Fs13318-022-00766-9/MediaObjects/13318_2022_766_MOESM1_ESM.pdf>

# Population

Data were pooled from four single-centre, single-dose, randomised crossover
studies in 116 healthy adults (Morse 2022 Section 2.1): MXIV-01 (n = 30) and
MXIV-06 (n = 30) compared intravenous acetaminophen and ibuprofen (alone or
combined, 15 min infusions; 30 min for the 400 mg ibuprofen product) with
combination tablets, fasted; MX-14a (n = 28) and MX-14b (n = 28) compared the
combination as film-coated tablets, ready-to-use oral suspension and a sachet
(powder dissolved in 200 mL water), fasted in MX-14a and fed in MX-14b. The
pooled participants (Table 1) were 18-49 years old (median 24), weighed
49-116 kg (median 70.8), were 156-199 cm tall (median 173) and 93 were male
and 23 female. The analysis used 6095 acetaminophen and 6046 ibuprofen
concentrations.

The same information is available programmatically through each model's
`population` metadata, e.g.
`readModelDb("Morse_2022_ibuprofen")()$population`.

# Model structure

Both drugs use a two-compartment disposition model with first-order
elimination. Oral doses enter a depot with bioavailability `F`, a lag time
`TLAG` and a first-order absorption rate `ka = ln(2) / T1/2ABS`; intravenous
doses are infusions into the central compartment.

*Size and body composition.* Fat-free mass is predicted from weight, height
and sex (Janmahasatian; Eq. 2). Normal fat mass is
`NFM = FFM + Ffat * (WT - FFM)` (Eqs. 3-4), and each parameter is scaled by
`(NFM / NFM_STD)^EXP` with `EXP` fixed at 3/4 for clearances and 1 for volumes
(Eqs. 1 and 5). The standard individual is a 70 kg, 1.76 m male whose FFM is
56.1 kg, so `NFM_STD = 56.1 + Ffat * (70 - 56.1)`. Acetaminophen uses
`Ffat = 0.816` for clearances and `Ffat = 1` (total body weight) for volumes;
ibuprofen uses 0.863 and 0.718.

*Formulation and food.* The fasted tablet is the reference for `T1/2ABS` and
`TLAG`. In the fasted state the suspension and the sachet each carry their own
factor; in the fed state each of the three oral formulations carries its own
factor, again relative to the fasted tablet (the fed factor replaces, rather
than multiplies, the fasted formulation factor). The section "Which reference
do the fed factors use?" below shows why.

*Residual error* is combined additive and proportional with a between-subject
random effect on its magnitude (Eq. 8):
`SD = sqrt((Cc * propSd)^2 + addSd^2) * exp(etaRUV)`.

# Source trace

Every `ini()` value carries an in-file source comment in
`inst/modeldb/specificDrugs/Morse_2022_acetaminophen.R` and
`inst/modeldb/specificDrugs/Morse_2022_ibuprofen.R`. They are collected here.

| Parameter | Acetaminophen | Ibuprofen | Source |
|-----------|---------------|-----------|--------|
| `lcl` (CL, L/h/70 kg) | 24.0 | 3.79 | Tables 2 / 3 |
| `lvc` (V1, L/70 kg) | 43.7 | 6.05 | Tables 2 / 3 |
| `lq` (Q2, L/h/70 kg) | 43.5 | 10.5 | Tables 2 / 3 |
| `lvp` (V2, L/70 kg) | 29.7 | 4.37 | Tables 2 / 3 |
| `lfdepot` (F) | 0.859 | 0.941 | Tables 2 / 3 (FPARA, FIBU) |
| `ltabs` (T1/2ABS, min) | 11.5 | 26.7 | Tables 2 / 3 |
| `ltlag` (TLAG, min) | 5.30 | 6.66 | Tables 2 / 3 |
| `ffat_cl` | 0.816 | 0.863 | Tables 2 / 3 (FFATCL) |
| `ffat_v` | 1 (fixed) | 0.718 | Tables 2 / 3 (FFATV) |
| `e_wt_cl`, `e_wt_vc` | 0.75, 1 (fixed) | 0.75, 1 (fixed) | Section 2.3 |
| `e_form_apapibu_susp_tabs` / `_tlag` | 0.394 / 0.743 | 0.719 / 0.984 | Tables 2 / 3 (F_FAST, suspension) |
| `e_form_powder_tabs` / `_tlag` | 0.462 / 0.845 | 0.235 / 0.539 | Tables 2 / 3 (F_FAST, sachet) |
| `e_fed_tabs_tablet` / `e_fed_tlag_tablet` | 1.87 / 4.63 | 1.59 / 3.65 | Tables 2 / 3 (F_FED, tablet) |
| `e_fed_tabs_susp` / `e_fed_tlag_susp` | 2.52 / 2.93 | 2.45 / 2.52 | Tables 2 / 3 (F_FED, suspension) |
| `e_fed_tabs_powder` / `e_fed_tlag_powder` | 2.30 / 2.10 | 3.79 / 0.178 | Tables 2 / 3 (F_FED, sachet) |
| IIV SD on CL, V1, Q2, V2 | 0.173, 0.537, 0.615, 0.466 | 0.221, 0.302, 0.660, 0.530 | Tables 2 / 3 (PPV%) |
| IIV correlations CL-V1-Q2-V2 | see model file | see model file | Supplementary Table S2 |
| IIV SD on F, T1/2ABS, TLAG | 0.145, 0.856, 0.960 | 0.051, 0.787, 0.809 | Tables 2 / 3 (PPV%) |
| `etaRUV` SD | 1.05 | 0.597 | Tables 2 / 3 (RUV PPV%) |
| `addSd` (mg/L) | 0.064 | 0.422 | Tables 2 / 3 (RUV ADD) |
| `propSd` | 0.070 | 0.240 | Tables 2 / 3 (RUV PROP) |
| FFM equation, WHS constants | -- | -- | Eq. 2 and Section 2.3 |
| NFM, NFM_STD, allometry | -- | -- | Eqs. 1, 3-5; Section 2.3 |
| Residual error with `etaRUV` | -- | -- | Eq. 8 |

The paper reports between-subject variability as `PPV% = sqrt(omega^2)`
(Section 2.4.1 and the table footnotes), so each variance is `(PPV/100)^2`.

# Which reference do the fed factors use?

The table footnotes state that the fasted tablet is the baseline for `T1/2ABS`
and `TLAG`, but a reader could still multiply a fed factor onto the fasted
formulation factor. The two readings give very different absorption for the
fed suspension and sachet, and Table 4's simulated `TMAX` separates them. The
check below solves the typical 70 kg, 1.76 m male under each reading. The
alternative reading is produced by overriding the fed factors with the product
of the fasted and fed factors.

```{r interpretation}
mod_apap <- readModelDb("Morse_2022_acetaminophen")
mod_ibu <- readModelDb("Morse_2022_ibuprofen")

scenarios <- tibble::tribble(
  ~formulation, ~FED,
  "Tablet", 0L,
  "Suspension", 0L,
  "Sachet", 0L,
  "Tablet", 1L,
  "Suspension", 1L,
  "Sachet", 1L
) |>
  mutate(
    scenario = paste(formulation, ifelse(FED == 1L, "fed", "fasted")),
    FORM_APAPIBU_SUSP = as.integer(formulation == "Suspension"),
    FORM_POWDER = as.integer(formulation == "Sachet")
  )

typical_events <- function(dose, tgrid = seq(0, 6, by = 1 / 120)) {
  doses <- scenarios |>
    mutate(id = row_number(), time = 0, amt = dose, evid = 1L, cmt = "depot")
  obs <- tidyr::crossing(doses |> select(-time, -amt, -evid, -cmt), time = tgrid) |>
    mutate(amt = 0, evid = 0L, cmt = "central")
  bind_rows(doses, obs) |>
    mutate(WT = 70, HT = 176, SEXF = 0L) |>
    arrange(id, time, desc(evid))
}

typical_tmax <- function(mod, dose, params = NULL) {
  ev <- typical_events(dose)
  s <- rxode2::rxSolve(rxode2::zeroRe(mod), ev,
    params = params,
    keep = "scenario", returnType = "data.frame"
  )
  s |>
    group_by(scenario) |>
    summarise(cmax = max(Cc), tmax = time[which.max(Cc)], .groups = "drop")
}

# Alternative: fed factor multiplied onto the fasted formulation factor
alt_params <- function(mod) {
  th <- rxode2::rxode(mod)$theta
  c(
    e_fed_tabs_susp = unname(th["e_fed_tabs_susp"] * th["e_form_apapibu_susp_tabs"]),
    e_fed_tlag_susp = unname(th["e_fed_tlag_susp"] * th["e_form_apapibu_susp_tlag"]),
    e_fed_tabs_powder = unname(th["e_fed_tabs_powder"] * th["e_form_powder_tabs"]),
    e_fed_tlag_powder = unname(th["e_fed_tlag_powder"] * th["e_form_powder_tlag"])
  )
}

table4 <- tibble::tribble(
  ~drug, ~scenario, ~cmax_med, ~cmax_p10, ~cmax_p90, ~tmax_med, ~tmax_p10, ~tmax_p90,
  "Acetaminophen", "Tablet fed", 10.9, 5.45, 19.8, 0.88, 0.42, 1.82,
  "Acetaminophen", "Suspension fed", 9.53, 4.81, 19.0, 1.10, 0.50, 2.17,
  "Acetaminophen", "Sachet fed", 9.91, 5.01, 19.4, 1.03, 0.51, 1.99,
  "Acetaminophen", "Tablet fasted", 12.8, 6.76, 24.2, 0.61, 0.26, 1.24,
  "Acetaminophen", "Suspension fasted", 15.08, 7.35, 31.2, 0.31, 0.13, 0.72,
  "Acetaminophen", "Sachet fasted", 14.4, 7.56, 28.6, 0.35, 0.16, 0.86,
  "Ibuprofen", "Tablet fed", 20.1, 10.3, 33.3, 1.2, 0.68, 2.17,
  "Ibuprofen", "Suspension fed", 16.2, 8.00, 30.3, 1.55, 0.82, 2.63,
  "Ibuprofen", "Sachet fed", 12.6, 6.14, 23.8, 1.90, 1.12, 3.21,
  "Ibuprofen", "Tablet fasted", 24.1, 13.3, 40.8, 0.94, 0.48, 1.67,
  "Ibuprofen", "Suspension fasted", 26.6, 14.8, 45.0, 0.77, 0.40, 1.50,
  "Ibuprofen", "Sachet fasted", 34.4, 20.8, 55.7, 0.38, 0.19, 0.80
)

interp <- bind_rows(
  full_join(
    typical_tmax(mod_apap, 1000) |> rename(cmax_packaged = cmax, tmax_packaged = tmax),
    typical_tmax(mod_apap, 1000, alt_params(mod_apap)) |>
      rename(cmax_alternative = cmax, tmax_alternative = tmax),
    by = "scenario"
  ) |> mutate(drug = "Acetaminophen"),
  full_join(
    typical_tmax(mod_ibu, 300) |> rename(cmax_packaged = cmax, tmax_packaged = tmax),
    typical_tmax(mod_ibu, 300, alt_params(mod_ibu)) |>
      rename(cmax_alternative = cmax, tmax_alternative = tmax),
    by = "scenario"
  ) |> mutate(drug = "Ibuprofen")
) |>
  left_join(table4 |> select(drug, scenario, cmax_med, tmax_med), by = c("drug", "scenario")) |>
  mutate(
    ratio_packaged = cmax_packaged / cmax_med,
    ratio_alternative = cmax_alternative / cmax_med
  )

interp |>
  select(drug, scenario, tmax_med, tmax_packaged, tmax_alternative, ratio_packaged, ratio_alternative) |>
  rename(
    "Drug" = drug,
    "Scenario" = scenario,
    "Table 4 median TMAX (h)" = tmax_med,
    "Typical TMAX, fed factor vs fasted tablet (h)" = tmax_packaged,
    "Typical TMAX, fed x fasted factors (h)" = tmax_alternative,
    "Typical CMAX / Table 4, fed factor vs fasted tablet" = ratio_packaged,
    "Typical CMAX / Table 4, fed x fasted factors" = ratio_alternative
  ) |>
  knitr::kable(digits = 2, caption = "Typical-value TMAX and CMAX (relative to the Table 4 median) under the two readings of the fed factors. The readings differ only for the fed suspension and fed sachet.")

# Deterministic (no random effects). Fasted rows do not depend on the
# reading, so their CMAX ratio is the yardstick for the fed rows.
fasted_ratio <- interp |>
  filter(grepl("fasted", scenario)) |>
  group_by(drug) |>
  summarise(yardstick = median(ratio_packaged), .groups = "drop")
fed_test <- interp |>
  filter(grepl("fed", scenario), !grepl("Tablet", scenario)) |>
  left_join(fasted_ratio, by = "drug")
stopifnot(
  # Summed TMAX error over the four discriminating rows
  sum(abs(fed_test$tmax_packaged - fed_test$tmax_med)) <
    sum(abs(fed_test$tmax_alternative - fed_test$tmax_med)),
  # Fed CMAX ratios stay close to the drug's fasted ratio only under the
  # packaged reading
  max(abs(fed_test$ratio_packaged - fed_test$yardstick)) <
    max(abs(fed_test$ratio_alternative - fed_test$yardstick))
)
```

The printed parameters do not reproduce the Table 4 `CMAX` medians exactly
under either reading (the next sections discuss why), but the fasted rows,
which do not depend on the reading, show how far off they are: a median of
about 0.93 of Table 4 for acetaminophen and 0.74 for ibuprofen at the
typical-value level. Under the packaged reading (each fed factor relative to
the fasted tablet) the fed suspension and sachet rows stay within 0.09 of
those ratios, and the summed `TMAX` error over the four rows is smaller. Under
the alternative the ibuprofen fed sachet would peak at 0.8 h with a `CMAX`
1.46 times the published median, whereas Table 4 shows it as the slowest and
lowest of all oral scenarios. The alternative is closer only for the
ibuprofen fed suspension `TMAX`, where the fasted suspension factors (0.719 and
0.984) are near 1 and the two readings barely differ. The packaged reading
also matches the abstract ("Feeding increased both absorption half-life and
absorption lag time when compared to the tablet formulation under fasting
conditions").

# Virtual cohort

Table 4 was simulated from the pooled demographics resampled with replacement.
The individual data are not public, so the virtual cohort draws sex, weight
and height to match Table 1: 23 of 116 female, log-normal weight around the
70.8 kg median truncated to the observed 49-116 kg, and normal height with
sex-specific means truncated to the observed 156-199 cm. Each of the six oral
scenarios has its own cohort of 200 participants, for each drug.

```{r cohort}
# set.seed() fixes the covariate draws below. rxode2's random effects use
# their own RNG (rxSetSeed), whose stream depends on the number of solver
# threads, so the simulated cohort differs between machines. Every assertion
# below is on a cohort median and has headroom for that.
set.seed(20220402)
rxode2::rxSetSeed(20220402)

n_per_arm <- 200L
clinical_grid <- c(5, 15, 30, 45) / 60
clinical_grid <- c(clinical_grid, 1, 1.25, 1.5, 2, 3, 6, 8, 10, 12)
fine_grid <- seq(0, 12, by = 1 / 12)

make_cohort <- function(i, dose, id_offset) {
  sc <- scenarios[i, ]
  sexf <- rbinom(n_per_arm, 1, 23 / 116)
  covs <- tibble(
    id = id_offset + seq_len(n_per_arm),
    SEXF = sexf,
    WT = pmin(pmax(exp(rnorm(n_per_arm, log(70.8), 0.17)), 49), 116),
    HT = pmin(pmax(rnorm(n_per_arm, ifelse(sexf == 1, 163, 175), 7), 156), 199),
    FED = sc$FED,
    FORM_APAPIBU_SUSP = sc$FORM_APAPIBU_SUSP,
    FORM_POWDER = sc$FORM_POWDER,
    scenario = sc$scenario
  )
  doses <- covs |> mutate(time = 0, amt = dose, evid = 1L, cmt = "depot")
  obs <- tidyr::crossing(covs, time = sort(unique(c(fine_grid, clinical_grid)))) |>
    mutate(amt = 0, evid = 0L, cmt = "central")
  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}

make_events <- function(dose) {
  bind_rows(lapply(seq_len(nrow(scenarios)), function(i) {
    make_cohort(i, dose, id_offset = (i - 1L) * n_per_arm)
  }))
}

events_apap <- make_events(1000)
events_ibu <- make_events(300)
stopifnot(!anyDuplicated(events_apap[, c("id", "time", "evid")]))
```

# Simulation

```{r simulate}
sim_apap <- rxode2::rxSolve(mod_apap, events_apap,
  keep = "scenario", returnType = "data.frame"
)
sim_ibu <- rxode2::rxSolve(mod_ibu, events_ibu,
  keep = "scenario", returnType = "data.frame"
)
sim_all <- bind_rows(
  sim_apap |> mutate(drug = "Acetaminophen"),
  sim_ibu |> mutate(drug = "Ibuprofen")
)
```

## Concentration-time profiles by formulation and food

```{r profiles, fig.width = 8, fig.height = 6}
sim_all |>
  filter(time <= 8) |>
  group_by(drug, scenario, time) |>
  summarise(
    Q10 = quantile(Cc, 0.10),
    Q50 = median(Cc),
    Q90 = quantile(Cc, 0.90),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.25) +
  geom_line() +
  facet_grid(drug ~ scenario, scales = "free_y") +
  labs(
    x = "Time after dose (h)", y = "Plasma concentration (mg/L)",
    caption = "Median and 10th-90th percentiles; acetaminophen 1000 mg, ibuprofen 300 mg."
  ) +
  theme(strip.text.x = element_text(size = 7))
```

# Replicate Table 4: simulated CMAX and TMAX

Table 4 reports medians and 10th / 90th percentiles of simulated `CMAX` and
`TMAX` after acetaminophen 1000 mg and ibuprofen 300 mg. Two summaries of the
virtual cohort are shown: the noise-free individual prediction on a 5 min grid,
and the observation with residual error read off the paper's sampling grid
(Section 2.7). The largest observation on a sparse grid overstates the true
peak when the residual error is large, which matters for ibuprofen
(proportional error 24% with 60% between-subject variability on its
magnitude).

```{r table4}
cmax_tmax <- function(d, value) {
  d |>
    group_by(drug, scenario, id) |>
    summarise(cmax = max(.data[[value]]), tmax = time[which.max(.data[[value]])], .groups = "drop") |>
    group_by(drug, scenario) |>
    summarise(
      cmax_med = median(cmax), cmax_p10 = quantile(cmax, 0.1), cmax_p90 = quantile(cmax, 0.9),
      tmax_med = median(tmax), tmax_p10 = quantile(tmax, 0.1), tmax_p90 = quantile(tmax, 0.9),
      .groups = "drop"
    )
}

on_clinical_grid <- sim_all |>
  filter(round(time * 60) %in% round(clinical_grid * 60))

sim_ipred <- cmax_tmax(sim_all, "Cc")
sim_obs <- cmax_tmax(on_clinical_grid, "sim")

fmt <- function(m, lo, hi, digits) {
  sprintf(paste0("%.", digits, "f (%.", digits, "f, %.", digits, "f)"), m, lo, hi)
}

table4_cmp <- table4 |>
  left_join(sim_ipred, by = c("drug", "scenario"), suffix = c("", "_ipred")) |>
  left_join(sim_obs, by = c("drug", "scenario"), suffix = c("", "_obs"))

table4_cmp |>
  transmute(
    drug, scenario,
    cmax_pub = fmt(cmax_med, cmax_p10, cmax_p90, 1),
    cmax_ipred = fmt(cmax_med_ipred, cmax_p10_ipred, cmax_p90_ipred, 1),
    cmax_obs = fmt(cmax_med_obs, cmax_p10_obs, cmax_p90_obs, 1),
    tmax_pub = fmt(tmax_med, tmax_p10, tmax_p90, 2),
    tmax_ipred = fmt(tmax_med_ipred, tmax_p10_ipred, tmax_p90_ipred, 2),
    tmax_obs = fmt(tmax_med_obs, tmax_p10_obs, tmax_p90_obs, 2)
  ) |>
  rename(
    "Drug" = drug,
    "Scenario" = scenario,
    "CMAX Table 4 (mg/L)" = cmax_pub,
    "CMAX simulated, IPRED (mg/L)" = cmax_ipred,
    "CMAX simulated, with RUV on sampling grid (mg/L)" = cmax_obs,
    "TMAX Table 4 (h)" = tmax_pub,
    "TMAX simulated, IPRED (h)" = tmax_ipred,
    "TMAX simulated, with RUV on sampling grid (h)" = tmax_obs
  ) |>
  knitr::kable(caption = "Median (10th, 90th percentile) CMAX and TMAX: Table 4 of Morse 2022 versus the packaged models.")
```

```{r table4-checks}
chk <- table4_cmp |>
  mutate(
    cmax_ratio_obs = cmax_med_obs / cmax_med,
    tmax_ratio_ipred = tmax_med_ipred / tmax_med
  )
knitr::kable(
  chk |>
    select(drug, scenario, cmax_ratio_obs, tmax_ratio_ipred) |>
    rename(
      "Drug" = drug,
      "Scenario" = scenario,
      "CMAX ratio, simulated with RUV / Table 4" = cmax_ratio_obs,
      "TMAX ratio, simulated IPRED / Table 4" = tmax_ratio_ipred
    ),
  digits = 2,
  caption = "Ratios of the simulated medians to the Table 4 medians."
)
# Medians over the 12 cells. The packaged models run about 15% below the
# Table 4 CMAX medians (see text); a mis-transcribed volume, bioavailability,
# dose or unit moves CMAX by a factor approaching 2 and leaves this band.
# TMAX: the fed scenarios run later than Table 4; a swapped fed-factor
# reference or a minutes-for-hours slip moves the median ratio far outside.
stopifnot(
  median(chk$cmax_ratio_obs) > 0.70,
  median(chk$cmax_ratio_obs) < 1.00,
  median(chk$tmax_ratio_ipred) > 0.85,
  median(chk$tmax_ratio_ipred) < 1.50
)
```

The simulated `CMAX` medians (with residual error, on the sampling grid) are
about 10-30% lower than Table 4, and the fed `TMAX` medians are 20-70% later;
the fasted acetaminophen `TMAX` medians agree to within 20%. The packaged parameters were not adjusted. Two observations from the
paper itself suggest that Table 4 does not follow directly from the printed
parameters:

- Even for the typical individual (no random effects), the printed parameters
  give an acetaminophen fasted-tablet `CMAX` of about 11.0 mg/L and an
  ibuprofen fasted-tablet `CMAX` of about 17.6 mg/L, below the Table 4
  medians of 12.8 and 24.1 mg/L.
- Table 4's upper percentiles exceed the observed concentrations in
  Supplementary Figures S3 and S4. The ibuprofen sachet 90th percentile is
  55.7 mg/L, while no observed sachet concentration (fasted and fed pooled)
  exceeds about 38 mg/L; the fasted ibuprofen suspension median is 26.6 mg/L,
  about the largest observed suspension concentration. The simulated
  percentiles from the packaged models sit inside the observed ranges.

The ordering of the scenarios (fasted sachet and suspension fastest; feeding
lowering `CMAX` and delaying `TMAX` for every formulation) is reproduced.

# PKNCA validation

PKNCA on the noise-free individual predictions, grouped by drug and scenario.
The paper reports no NCA of its own beyond Table 4, so the PKNCA medians are
compared with the Table 4 medians.

```{r pknca}
sim_nca <- sim_all |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, drug, scenario)
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, drug, scenario) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, drug, scenario, time, .keep_all = TRUE) |>
  arrange(drug, scenario, id, time)

dose_nca <- bind_rows(
  events_apap |> filter(evid == 1) |> mutate(drug = "Acetaminophen"),
  events_ibu |> filter(evid == 1) |> mutate(drug = "Ibuprofen")
) |>
  select(id, time, amt, drug, scenario)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | drug + scenario + id)
dose_obj <- PKNCA::PKNCAdose(dose_nca, amt ~ time | drug + scenario + id)
intervals <- data.frame(
  start = 0, end = 12,
  cmax = TRUE, tmax = TRUE, auclast = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

reference <- table4 |>
  transmute(drug, scenario, cmax = cmax_med, tmax = tmax_med)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = reference,
  by = c("drug", "scenario"),
  units = c(cmax = "mg/L", tmax = "h"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "PKNCA medians of the noise-free simulations versus the Table 4 medians. * differs from the reference by more than 20%.")
```

The starred rows are the `CMAX` and fed `TMAX` differences discussed in the
previous section. The noise-free `CMAX` medians are lower than the
residual-error summaries above, because the largest of several noisy
observations overstates the true peak. Table 4's footnote says `CMAX` and
`TMAX` were calculated from simulations; it does not say whether residual error
was included.

# Total body composition: typical clearance and volume across weight

```{r size, fig.width = 7, fig.height = 3.5}
size_grid <- tidyr::crossing(WT = seq(50, 120, by = 5), SEXF = c(0L, 1L)) |>
  mutate(id = row_number(), HT = ifelse(SEXF == 1L, 163, 176), time = 0, amt = 0, evid = 0L, cmt = "central",
         FED = 0L, FORM_APAPIBU_SUSP = 0L, FORM_POWDER = 0L)
size_par <- bind_rows(
  rxode2::rxSolve(rxode2::zeroRe(mod_apap), size_grid, keep = c("WT", "SEXF"), returnType = "data.frame") |>
    mutate(drug = "Acetaminophen"),
  rxode2::rxSolve(rxode2::zeroRe(mod_ibu), size_grid, keep = c("WT", "SEXF"), returnType = "data.frame") |>
    mutate(drug = "Ibuprofen")
) |>
  select(drug, WT, SEXF, cl, vc) |>
  tidyr::pivot_longer(c(cl, vc), names_to = "parameter") |>
  mutate(parameter = ifelse(parameter == "cl", "CL (L/h)", "V1 (L)"), sex = ifelse(SEXF == 1, "female, 163 cm", "male, 176 cm"))

ggplot(size_par, aes(WT, value, colour = sex)) +
  geom_line() +
  facet_wrap(drug ~ parameter, scales = "free_y", nrow = 1) +
  labs(x = "Total body weight (kg)", y = "Typical value", colour = NULL) +
  theme(legend.position = "bottom")
```

# Pain-score simulation (Supplementary Table S4)

Morse 2022 simulated the time to a 2-point reduction in a 0-10 pain score for
a typical 70 kg individual, using the PK models above and the
acetaminophen + ibuprofen pharmacodynamic model of Hannam 2018
(*Paediatr Anaesth* 28:841-851, reproduced as Supplementary Table S3). The
pharmacodynamic model belongs to Hannam 2018 and is not part of the packaged
models; it is coded here only to check the absorption parameters against
Supplementary Table S4. Each drug is simulated alone with an effect
compartment and a sigmoid Emax model, `E = EMAX * Ce^HILL / (C50^HILL + Ce^HILL)`,
and the pain-score reduction is taken as `10 * E` (a baseline of 10 is an
assumption; the supplement does not state the baseline score).

```{r pain}
# Hannam 2018 parameters (Supplementary Table S3) enter as data columns:
# tkeo = equilibration half-time (h), c50 (mg/L); EMAX 0.648 and HILL 1.48
# are shared by both drugs.
pd_layer <- function(mod) {
  mod |>
    rxode2::zeroRe() |>
    rxode2::model(d / dt(ce) <- log(2) / tkeo * (Cc - ce), append = TRUE) |>
    rxode2::model(pain_red <- 10 * 0.648 * ce^1.48 / (c50^1.48 + ce^1.48), append = TRUE)
}

time_to_2 <- function(mod, dose, tkeo, c50, params = NULL) {
  ev <- typical_events(dose, tgrid = seq(0, 3, by = 1 / 600)) |>
    filter(FED == 0L) |>
    mutate(tkeo = tkeo, c50 = c50)
  s <- rxode2::rxSolve(mod, ev, params = params, keep = "scenario", returnType = "data.frame")
  s |>
    group_by(scenario) |>
    summarise(minutes = 60 * time[which(pain_red >= 2)[1]], .groups = "drop")
}

pd_apap <- pd_layer(mod_apap)
pd_ibu <- pd_layer(mod_ibu)
no_lag <- c(ltlag = log(1e-6))

pain <- bind_rows(
  full_join(time_to_2(pd_apap, 1000, 0.34, 7.06) |> rename(with_lag = minutes),
    time_to_2(pd_apap, 1000, 0.34, 7.06, no_lag) |> rename(no_lag = minutes),
    by = "scenario"
  ) |> mutate(drug = "Acetaminophen"),
  full_join(time_to_2(pd_ibu, 300, 1.04, 3.95) |> rename(with_lag = minutes),
    time_to_2(pd_ibu, 300, 1.04, 3.95, no_lag) |> rename(no_lag = minutes),
    by = "scenario"
  ) |> mutate(drug = "Ibuprofen")
) |>
  left_join(
    tibble::tribble(
      ~drug, ~scenario, ~published,
      "Acetaminophen", "Tablet fasted", 21.8,
      "Acetaminophen", "Suspension fasted", 14.3,
      "Acetaminophen", "Sachet fasted", 15.0,
      "Ibuprofen", "Tablet fasted", 24.0,
      "Ibuprofen", "Suspension fasted", 24.8,
      "Ibuprofen", "Sachet fasted", 14.3
    ),
    by = c("drug", "scenario")
  )

pain |>
  select(drug, scenario, published, with_lag, no_lag) |>
  rename(
    "Drug" = drug,
    "Scenario" = scenario,
    "Table S4 (min)" = published,
    "Simulated, with lag time (min)" = with_lag,
    "Simulated, lag time removed (min)" = no_lag
  ) |>
  knitr::kable(digits = 1, caption = "Time to a 2-point pain-score reduction for a typical 70 kg individual, fasted.")

# Deterministic typical-value checks: with the lag time every time is within
# 10 min of Supplementary Table S4; without it the acetaminophen times are
# within 1 min.
acet_no_lag <- pain |> filter(drug == "Acetaminophen")
stopifnot(
  all(abs(pain$with_lag - pain$published) < 10),
  all(abs(acet_no_lag$no_lag - acet_no_lag$published) < 1)
)
```

The acetaminophen times in Supplementary Table S4 are reproduced to within
0.2 min when the lag time is removed from the simulation, and are about 4-5
min later with the lag time. This suggests the published pain-score
simulations omitted the absorption lag time for acetaminophen. The ibuprofen
tablet is also reproduced (to within 0.5 min) without the lag time, but the ibuprofen suspension and sachet
are not reproduced by either variant; the source of that difference could not
be identified from the paper.

# Assumptions and deviations

- **Scaling of Q2 and V2.** The paper gives the size and body-composition
  model for "clearance" and "volume of distribution" (Eqs. 1-5) and reports Q2
  in L/h/70 kg and V2 in L/70 kg. The packaged models apply the clearance size
  factor (Ffat for CL, exponent 3/4) to Q2 and the volume size factor (Ffat for
  V, exponent 1) to V2, which is the usual convention of this group of authors.
- **NFM standard value.** `NFM_STD = 56.1 + Ffat * (70 - 56.1)` follows the
  paper's definition (a 70 kg, 1.76 m male with FFM 56.1 kg). The packaged
  Janmahasatian equation gives 56.1 kg for that individual.
- **Random effect on bioavailability.** Eq. 6 states that every random effect
  is exponential, so `F = F_pop * exp(eta)`. With the ibuprofen `F = 0.941`
  and a 5.1% PPV, about one subject in eight has an individual `F` slightly
  above 1. The paper does not report a logit or capped form.
- **Residual error.** Eq. 8 writes the proportional term on the observation
  (`Obs`); it is implemented on the model prediction, as NONMEM residual
  error models are. The additive and proportional SDs are combined as a sum of
  variances, as in Eq. 8, and both are scaled by `exp(etaRUV)`.
- **Tablet strengths.** The 500/150 mg and 325/97.5 mg film-coated tablets are
  modelled as one tablet formulation, as in the paper.
- **Abstract values.** The abstract gives the acetaminophen and ibuprofen
  central volumes as 43.5 and 10.5 L/70 kg; those are the Q2 values of Tables
  2 and 3. The packaged models use the table values (V1 43.7 and 6.05 L/70 kg).
  The abstract and Discussion also give the acetaminophen bioavailability as
  86% and 87%; Table 2 gives 0.859, which is used.
- **Ibuprofen RUV PROP.** Table 3 prints 24.0% with a bootstrap 95% CI of
  19.9-22.1%, which excludes the estimate; the point estimate is used.
- **Virtual cohort.** Weight, height and sex are drawn to match Table 1 (no
  individual demographics are published). Ages are not needed by the models.
- **Pain-score model.** The Hannam 2018 pharmacodynamic layer is used only in
  this article, with an assumed baseline pain score of 10, to check
  Supplementary Table S4. It is not part of the packaged models.
- **Errata.** No correction notice for this article was found in Europe PMC
  (checked 2026-10-02).
