---
title: "Aripiprazole (Mou 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aripiprazole (Mou 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Mou_2024_aripiprazole"))$reference`
* Description: `r rxode2::rxode(readModelDb("Mou_2024_aripiprazole"))$description`
* Article: <https://doi.org/10.1177/20420986241303432>
* Supplement: supplemental material accompanying the article (virtual-population
  demographics Table S1, validation-study list Table S2, the metabolic-pathway
  derivation in section 1.3, and the dose-optimization Table S3)

# What this model is, and what it is not

Mou 2024 is a whole-body physiologically based pharmacokinetic (PBPK) analysis
built in PK-Sim 10.0. It has no published disposition ODEs, no volume of
distribution, and no half-life: aripiprazole distribution in the source is a
PK-Sim partition-coefficient database output, not a reported parameter.

What the paper *does* publish, completely, is the **clearance layer**:

* a systemic plasma clearance (Table 1 and supplemental section 1.3, stated two
  independent ways that agree to 0.1%),
* the split of that clearance between CYP2D6 and CYP3A4 and between the
  dehydrogenation and non-dehydrogenation routes (Table 2 enzyme kinetics plus
  Table 3 enzyme abundances),
* how each of those changes with CYP2D6 phenotype (Table 2), with Child-Pugh A
  hepatic impairment, and with risperidone coadministration (Table 5,
  supplemental Table S3),
* and complete simulated concentration-time profiles for both analytes across
  four CYP2D6 phenotypes (Figure 2) and for mild hepatic impairment (Figure 4).

The packaged model is therefore a **compartmental reduction**, not a
reimplementation of the PBPK model. Every clearance and every covariate effect
is taken from the paper. The five distribution parameters and the absorption
rate constant, which the paper does not report at all, were recovered by
digitizing the predicted mean curves of Figure 2(a) and fitting the reduction to
them with all clearances held fixed. The reduction is then checked against
published summaries that were **not** used in the fit.

# Population

The simulations use a PK-Sim virtual population of 500 individuals
(supplemental Table S1): age 18-65 years (median 43), weight 48.0-171.2 kg
(median 78.0), height 151.8-191.7 cm (median 170.8). Where a source trial did
not report demographics, the authors assumed a 100% Mexican-American / White
male population aged 18-65 and completed the covariates from NHANES.

Model validation used four clinical studies (supplemental Table S2):

| Model | Dosing | n | Age (years) | Source |
|---|---|---|---|---|
| DGI (NM / IM / PM / UM) | 10 mg single | 28 / 21 / 1 / 5 | 19.0-51.0 | Kneller 2021 |
| Initial | 10 mg multiple | 8 | 18.0-60.0 | Mallikaarjun 2004 |
| Initial | 30 mg multiple | 7 | 23.0-35.6 | Mallikaarjun 2004 |
| DDI | 20 mg multiple + risperidone | 10 | 21.0-77.0 | Waade 2009 |
| DDZI | 15 mg single, Child-Pugh A | 8 | 43.0-61.0 | Mallikaarjun 2008 |

The same information is available programmatically via
`readModelDb("Mou_2024_aripiprazole")()$population`.

# Source trace

Every `ini()` entry, with the location it came from.

| Parameter | Value | Source |
|---|---|---|
| `lcl_nonren` | 47.96 mL/h/kg (3.3572 L/h at 70 kg) | Table 1 `CL_H`, aripiprazole. Supplemental 1.3 independently states "plasma clearance 0.8 mL/min/kg" = 48.0 mL/h/kg |
| `lcl_renal` | 0.04 mL/h/kg | Table 1 `CL_R`, aripiprazole; supplemental 1.3 "renal clearance < 1%" |
| `lcl_nonren_dehydro` | 40.00 mL/h/kg | Table 1 `CL_H`, dehydro-aripiprazole |
| `lcl_renal_dehydro` | 0.06 mL/h/kg | Table 1 `CL_R`, dehydro-aripiprazole |
| `fmet2d6Dehydro` | 0.09017 | Table 2 `Kcat,CYP2D6,dehydrogenation` 14.99 /min, `Km` 26.20 uM; Table 3 `[CYP2D6]` 0.40 uM |
| `fmet2d6Other` | 0.34387 | Table 2 `Kcat,CYP2D6,hydroxylation` 57.17 /min |
| `fmet3a4Dehydro` | 0.31327 | Table 2 `Kcat,CYP3A4,dehydrogenation` 54.85 /min, `Km` 298.00 uM; Table 3 `[CYP3A4]` 4.32 uM |
| `fmet3a4Other` | 0.25268 | Table 2 `Kcat,CYP3A4,other` 44.24 /min |
| `rel2d6Im` | 0.6999 | Table 2 IM column: 10.49/14.99 and 40.02/57.17 |
| `rel3a4PmDehydro` | 1.3453 | Table 2 PM column footnote a: 73.79/54.85 |
| `rel3a4PmOther` | 1.3422 | Table 2 PM column footnote a: 59.38/44.24 |
| `mwRatioDehydro` | 0.99554 | Table 1 molecular weights 446.4 / 448.4 g/mol |
| `e_hepimp_cl` | 0.9279 | Supplemental Table S3: DDZI AUC 806.39 vs DGI NM 748.25 umol*min/L |
| `e_risp_cl` | 0.9907 | Supplemental Table S3: DDI AUC 755.26 vs DGI NM 748.25 umol*min/L |
| `lfdepot` | 0.974 | Derived: reconciles Table 1 `CL_H` at 70 kg with the reported single-dose AUC of 773.85 umol*min/L after 20 mg |
| `lka`, `lvc`, `lvp`, `lq` | 2.3638 /h, 190.8 L, 32.0 L, 4.629 L/h | Digitized from Figure 2(a); not reported by Mou 2024 |
| `lvc_dehydro`, `lvp_dehydro`, `lq_dehydro` | 43.2 L, 193.3 L, 9.952 L/h | Digitized from Figure 2(a); not reported by Mou 2024 |
| `propSd`, `propSd_dehydro` | 0.10 (placeholder) | Not reported; Mou 2024 evaluate the model by MPE / MAPE, not by a fitted sigma |

## The clearance layer is internally over-determined

Three statements in the paper independently confirm the way Table 2 was read
here. Each is reproduced by the packaged `ini()` values without adjustment.

```{r pathway-check}
E2D6 <- 0.40; E3A4 <- 4.32          # Table 3, normal liver (umol/L)
Km2D6 <- 26.20; Km3A4 <- 298.00     # Table 2 (umol)
clu <- function(kcat, E, Km) kcat * E / Km

d2 <- clu(14.99, E2D6, Km2D6); h2 <- clu(57.17, E2D6, Km2D6)
d3 <- clu(54.85, E3A4, Km3A4); o3 <- clu(44.24, E3A4, Km3A4)
tot <- d2 + h2 + d3 + o3

tibble::tibble(
  Quantity = c("CYP2D6 fraction of metabolism",
               "CYP3A4 fraction of metabolism",
               "Fraction forming dehydro-aripiprazole",
               "Aripiprazole plasma clearance (mL/min/kg)"),
  `From Table 1 / 2 / 3` = c(
    sprintf("%.4f", (d2 + h2) / tot),
    sprintf("%.4f", (d3 + o3) / tot),
    sprintf("%.4f", (d2 + d3) / tot),
    sprintf("%.4f", 47.96 / 60)),
  `Stated by the paper` = c("0.43 (suppl. 1.3)", "0.56 (suppl. 1.3)",
                            "0.40 (suppl. 1.3)", "0.8 (suppl. 1.3)")
) |>
  knitr::kable(caption = "Table 2 + Table 3 reproduce the three metabolic fractions and the clearance that the supplement states in prose.")
```

# Simulation setup

The model carries no between-subject variability: Mou 2024 report a PBPK
forward prediction with a virtual-population confidence band, not a population
model with estimated omegas. Simulations below are therefore typical-value
(deterministic) profiles for a 70 kg adult.

```{r sim-helper}
mod <- rxode2::rxode(readModelDb("Mou_2024_aripiprazole"))

# The model has two endpoints (Cc, Cc_dehydro), so observation rows carry
# dvid = 1 with cmt = NA; the algebraic observables are returned as columns.
mou_sim <- function(dose = 10, phenotype = "NM", hepimp = 0, risperidone = 0,
                    wt = 70, tmax = 1200, dt = 0.25) {
  grid <- seq(0, tmax, by = dt)
  ev <- dplyr::bind_rows(
    data.frame(id = 1L, time = 0, amt = dose, evid = 1L,
               cmt = "depot", dvid = NA_integer_),
    data.frame(id = 1L, time = grid, amt = NA_real_, evid = 0L,
               cmt = NA_character_, dvid = 1L)
  )
  ev$WT <- wt
  ev$CYP2D6_EM <- as.integer(phenotype == "NM")
  ev$CYP2D6_PM <- as.integer(phenotype == "PM")
  ev$CYP2D6_UM <- as.integer(phenotype == "UM")
  ev$HEPIMP_MILD <- hepimp
  ev$CONMED_RISPERIDONE <- risperidone
  out <- rxode2::rxSolve(mod, ev, returnType = "data.frame")
  out$phenotype <- phenotype
  out
}

trapz <- function(time, conc) sum(diff(time) * (utils::head(conc, -1) + utils::tail(conc, -1)) / 2)
```

# Replicating Figure 2: CYP2D6 phenotypes, single 10 mg dose

Mou 2024 Figure 2 plots the predicted mean concentration-time curves for
aripiprazole and dehydro-aripiprazole in each of the four CYP2D6 phenotypes.
The points below were digitized from the published mean curves of panel (a),
CYP2D6 normal metabolizer -- these are the curves the distribution parameters
were fitted to, so this panel is a fit check rather than an independent test.

```{r fig2a-digitized}
fig2a <- tibble::tribble(
  ~analyte,       ~time, ~conc,
  "aripiprazole",   2.8, 47.156, "aripiprazole",   4.0, 45.150,
  "aripiprazole",   5.0, 43.536, "aripiprazole",   6.4, 41.824,
  "aripiprazole",   7.2, 40.894, "aripiprazole",   9.2, 39.085,
  "aripiprazole",  11.8, 37.030, "aripiprazole",  16.0, 34.193,
  "aripiprazole",  20.0, 31.845, "aripiprazole",  24.8, 29.595,
  "aripiprazole",  30.0, 27.051, "aripiprazole",  36.0, 24.507,
  "aripiprazole",  42.0, 22.404, "aripiprazole",  48.6, 20.203,
  "aripiprazole",  54.0, 18.735, "aripiprazole",  60.0, 17.023,
  "aripiprazole",  63.4, 16.289,
  "dehydro-aripiprazole",  1.8, 1.810, "dehydro-aripiprazole",  9.0, 4.158,
  "dehydro-aripiprazole", 11.2, 4.500, "dehydro-aripiprazole", 16.0, 5.283,
  "dehydro-aripiprazole", 20.0, 5.577, "dehydro-aripiprazole", 23.2, 5.821,
  "dehydro-aripiprazole", 30.0, 6.164, "dehydro-aripiprazole", 36.0, 6.310,
  "dehydro-aripiprazole", 42.0, 6.408, "dehydro-aripiprazole", 47.0, 6.359,
  "dehydro-aripiprazole", 54.0, 6.310, "dehydro-aripiprazole", 60.0, 6.164,
  "dehydro-aripiprazole", 63.6, 6.066, "dehydro-aripiprazole", 69.0, 5.968
)

sim_nm <- mou_sim(dose = 10, phenotype = "NM", tmax = 75, dt = 0.1)
pred_nm <- dplyr::bind_rows(
  data.frame(analyte = "aripiprazole", time = sim_nm$time, conc = sim_nm$Cc),
  data.frame(analyte = "dehydro-aripiprazole", time = sim_nm$time, conc = sim_nm$Cc_dehydro)
)

ggplot2::ggplot(pred_nm, ggplot2::aes(time, conc, colour = analyte)) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::geom_point(data = fig2a, size = 2.2, shape = 21, fill = "white") +
  ggplot2::labs(
    x = "Time (h)", y = "Plasma concentration (ng/mL)", colour = NULL,
    title = "Replicates Figure 2(a) of Mou 2024",
    subtitle = "Lines: packaged model. Points: digitized published mean curves. Oral 10 mg, CYP2D6 NM."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

```{r fig2a-fit-quality}
interp <- function(sim, col, times) stats::approx(sim$time, sim[[col]], times)$y
fit_tbl <- fig2a |>
  dplyr::mutate(
    predicted = ifelse(analyte == "aripiprazole",
                       interp(sim_nm, "Cc", time),
                       interp(sim_nm, "Cc_dehydro", time))
  ) |>
  dplyr::group_by(analyte) |>
  dplyr::summarise(
    n = dplyr::n(),
    `RMSE (log scale)` = sqrt(mean((log(predicted) - log(conc))^2)),
    `Max |deviation| (%)` = max(abs(100 * (predicted / conc - 1))),
    .groups = "drop"
  ) |>
  dplyr::rename(Analyte = analyte, `N points` = n)

knitr::kable(fit_tbl, digits = 3,
             caption = "Agreement between the packaged model and the digitized Figure 2(a) mean curves.")
```

# Independent checks that were not used in the fit

## Published AUC after a single 20 mg dose

The Results section states that after a single 20 mg dose the aripiprazole AUC
was 773.85 umol*min/L. Nothing in that number entered the digitization fit
except as a weak anchor on the 10 mg curve; the 20 mg simulation is a scale
check.

```{r auc-20mg}
s20 <- mou_sim(dose = 20, phenotype = "NM")
auc20_umol_min <- trapz(s20$time, s20$Cc) / 448.4 * 60   # ng*h/mL -> umol*min/L

tibble::tibble(
  Quantity = "AUC after a single oral 20 mg dose (umol*min/L)",
  Simulated = round(auc20_umol_min, 2),
  Published = 773.85,
  `Difference (%)` = round(100 * (auc20_umol_min / 773.85 - 1), 1)
) |>
  knitr::kable(caption = "Mou 2024 Results, DDI models section.")
```

## Figure 2(c): the poor-metabolizer panel exposes two parameterizations

Mou 2024 report two different CYP2D6 poor-metabolizer results that do not agree
with each other:

* the Results text and abstract give a PM/NM AUC fold change of **1.75**;
* supplemental Table S3 and Table 5, which drive the dose recommendations, give
  1025.64/748.25 = **1.371**.

The difference is exactly the Table 2 footnote-a CYP3A4 up-adjustment. Setting
CYP2D6 activity to zero and leaving CYP3A4 at its normal-metabolizer values
gives an AUC ratio of 1.77 and reproduces the published Figure 2(c) curve to
about 2%; applying the footnote-a values as well gives 1.31, which is the Table
S3 value. The packaged model encodes Table 2 **as printed** (footnote a
included), because that is the parameterization behind the paper's dose
recommendations.

```{r fig2c}
fig2c <- tibble::tribble(
  ~time, ~conc,
   3.0, 51.302,  4.0, 49.363,  5.0, 47.535,  6.0, 46.316,
   7.2, 44.875,  9.8, 42.659, 12.0, 41.163, 16.0, 39.114,
  20.0, 37.507, 24.0, 36.011, 30.0, 34.072, 36.0, 32.244,
  42.0, 30.582, 47.2, 29.197, 54.0, 27.424, 60.0, 25.873,
  66.0, 24.488
)

sim_pm <- mou_sim(dose = 10, phenotype = "PM", tmax = 75, dt = 0.1)

# The Figure-2 variant: CYP2D6 zeroed, CYP3A4 left at normal-metabolizer values.
mod_fig2 <- rxode2::rxode(readModelDb("Mou_2024_aripiprazole"))
sim_pm_fig2 <- local({
  grid <- seq(0, 75, by = 0.1)
  ev <- dplyr::bind_rows(
    data.frame(id = 1L, time = 0, amt = 10, evid = 1L, cmt = "depot", dvid = NA_integer_),
    data.frame(id = 1L, time = grid, amt = NA_real_, evid = 0L, cmt = NA_character_, dvid = 1L))
  ev$WT <- 70; ev$CYP2D6_EM <- 0L; ev$CYP2D6_PM <- 1L; ev$CYP2D6_UM <- 0L
  ev$HEPIMP_MILD <- 0; ev$CONMED_RISPERIDONE <- 0
  rxode2::rxSolve(mod_fig2, ev, params = c(rel3a4PmDehydro = 1, rel3a4PmOther = 1),
                  returnType = "data.frame")
})

ggplot2::ggplot() +
  ggplot2::geom_line(data = data.frame(time = sim_pm$time, conc = sim_pm$Cc,
                                       src = "Packaged model (Table 2 as printed)"),
                     ggplot2::aes(time, conc, colour = src), linewidth = 0.9) +
  ggplot2::geom_line(data = data.frame(time = sim_pm_fig2$time, conc = sim_pm_fig2$Cc,
                                       src = "CYP2D6 zeroed only (Figure 2 variant)"),
                     ggplot2::aes(time, conc, colour = src), linewidth = 0.9, linetype = 2) +
  ggplot2::geom_point(data = fig2c, ggplot2::aes(time, conc), size = 2.2,
                      shape = 21, fill = "white") +
  ggplot2::labs(x = "Time (h)", y = "Aripiprazole (ng/mL)", colour = NULL,
                title = "Replicates Figure 2(c) of Mou 2024 (CYP2D6 poor metabolizer)",
                subtitle = "Points: digitized published mean curve. Oral 10 mg.") +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

```{r fig2c-quality}
tibble::tibble(
  Parameterization = c("Table 2 as printed (footnote a CYP3A4)",
                       "CYP2D6 zeroed only"),
  `RMSE vs Figure 2(c) (log scale)` = c(
    sqrt(mean((log(interp(sim_pm, "Cc", fig2c$time)) - log(fig2c$conc))^2)),
    sqrt(mean((log(interp(sim_pm_fig2, "Cc", fig2c$time)) - log(fig2c$conc))^2))),
  `PM/NM AUC ratio` = c(
    trapz(mou_sim(10, "PM")$time, mou_sim(10, "PM")$Cc) /
      trapz(mou_sim(10, "NM")$time, mou_sim(10, "NM")$Cc),
    1.767),
  `Paper value matched` = c("1.371 (Table S3, Table 5)", "1.75 (Results, abstract)")
) |>
  knitr::kable(digits = 3,
               caption = "The two CYP2D6 poor-metabolizer parameterizations present in Mou 2024.")
```

# Reproducing Figure 4: mild hepatic impairment, single 15 mg dose

```{r fig4}
sim_hep <- mou_sim(dose = 15, phenotype = "NM", hepimp = 1, tmax = 75, dt = 0.1)
sim_norm15 <- mou_sim(dose = 15, phenotype = "NM", tmax = 75, dt = 0.1)

dplyr::bind_rows(
  data.frame(time = sim_hep$time, conc = sim_hep$Cc,
             grp = "Child-Pugh A, aripiprazole"),
  data.frame(time = sim_hep$time, conc = sim_hep$Cc_dehydro,
             grp = "Child-Pugh A, dehydro-aripiprazole"),
  data.frame(time = sim_norm15$time, conc = sim_norm15$Cc,
             grp = "Normal liver, aripiprazole")
) |>
  ggplot2::ggplot(ggplot2::aes(time, conc, colour = grp, linetype = grp)) +
  ggplot2::geom_line(linewidth = 0.9) +
  ggplot2::labs(x = "Time (h)", y = "Plasma concentration (ng/mL)",
                colour = NULL, linetype = NULL,
                title = "Replicates Figure 4 of Mou 2024",
                subtitle = "Oral single 15 mg aripiprazole, adults with mild hepatic impairment.") +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

# PKNCA validation

Non-compartmental analysis of the simulated profiles, run across the six
scenarios the paper reports an AUC for. All scenarios use a single oral 20 mg
dose so they are directly comparable with the paper's `umol*min/L` AUC table.

```{r pknca}
scenarios <- tibble::tribble(
  ~scenario,                 ~phenotype, ~hepimp, ~risperidone,
  "CYP2D6 NM",               "NM",       0,       0,
  "CYP2D6 IM",               "IM",       0,       0,
  "CYP2D6 PM",               "PM",       0,       0,
  "CYP2D6 UM",               "UM",       0,       0,
  "Child-Pugh A (NM)",       "NM",       1,       0,
  "With risperidone (NM)",   "NM",       0,       1
)

sim_all <- do.call(rbind, lapply(seq_len(nrow(scenarios)), function(i) {
  s <- mou_sim(dose = 20, phenotype = scenarios$phenotype[i],
               hepimp = scenarios$hepimp[i], risperidone = scenarios$risperidone[i],
               tmax = 600, dt = 0.5)
  data.frame(id = 1L, scenario = scenarios$scenario[i], time = s$time, Cc = s$Cc)
}))

# Only !is.na(Cc): keeping the time-zero record is required by PKNCA.
conc_data <- sim_all |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, scenario, time, Cc)

dose_data <- conc_data |>
  dplyr::distinct(id, scenario) |>
  dplyr::mutate(time = 0, dose = 20)

o_conc <- PKNCA::PKNCAconc(conc_data, Cc ~ time | id / scenario)
o_dose <- PKNCA::PKNCAdose(dose_data, dose ~ time | id / scenario)
o_data <- PKNCA::PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(start = 0, end = Inf,
                         cmax = TRUE, tmax = TRUE,
                         aucinf.obs = TRUE, half.life = TRUE)
)
res <- suppressWarnings(PKNCA::pk.nca(o_data))

nca_res <- as.data.frame(res) |>
  dplyr::select(scenario, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

nca_res |>
  dplyr::rename(Scenario = scenario, `Cmax (ng/mL)` = cmax, `Tmax (h)` = tmax,
                `AUC0-inf (ng*h/mL)` = aucinf.obs, `t1/2 (h)` = half.life) |>
  knitr::kable(digits = c(0, 2, 1, 0, 1),
               caption = "Simulated NCA parameters, single oral 20 mg aripiprazole.")
```

## Comparison against the published exposures

Mou 2024 reports AUC in umol*min/L; the reference column below converts those
to ng*h/mL using the aripiprazole molecular weight of 448.4 g/mol from Table 1.
The Child-Pugh A and risperidone rows come from supplemental Table S3, the four
phenotype rows from Table S3 as well.

```{r nca-compare}
published <- tibble::tibble(
  scenario = c("CYP2D6 NM", "CYP2D6 IM", "CYP2D6 PM", "CYP2D6 UM",
               "Child-Pugh A (NM)", "With risperidone (NM)"),
  aucinf.obs = c(748.25, 852.47, 1025.64, 748.25, 806.39, 755.26) / 60 * 448.4
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "scenario",
  params        = "aucinf.obs",
  units         = c(aucinf.obs = "ng*h/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  digits = 0,
  caption = "Simulated versus published aripiprazole AUC after a single oral 20 mg dose, by scenario. * marks rows differing by more than 20%."
)
```

# Reproducing the dose-optimization table

Mou 2024 converts each exposure change into a correction factor
`CF = AUC_reference / AUC_scenario` and a recommended dose change
`dD = 1 - CF` (equation 9 and Table 5 / Table S3).

```{r dose-optimization}
auc_of <- function(ph, hp = 0, rp = 0) {
  s <- mou_sim(dose = 20, phenotype = ph, hepimp = hp, risperidone = rp)
  trapz(s$time, s$Cc)
}
ref <- auc_of("NM")

dose_tbl <- tibble::tribble(
  ~Scenario,                              ~ph,  ~hp, ~rp, ~cf_paper, ~dd_paper,
  "CYP2D6 IM",                            "IM",  0,   0,   0.88,      "12%",
  "CYP2D6 PM",                            "PM",  0,   0,   0.73,      "27%",
  "CYP2D6 UM",                            "UM",  0,   0,   1.00,      "0%",
  "Child-Pugh A",                         "NM",  1,   0,   0.93,      "7%",
  "Risperidone coadministration",         "NM",  0,   1,   0.99,      "1%"
) |>
  dplyr::rowwise() |>
  dplyr::mutate(cf_model = ref / auc_of(ph, hp, rp)) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    `Correction factor (model)` = round(cf_model, 2),
    `Dose change (model)` = sprintf("%.0f%%", 100 * (1 - cf_model))
  ) |>
  dplyr::select(Scenario,
                `Correction factor (model)`, `Dose change (model)`,
                `Correction factor (paper)` = cf_paper,
                `Dose change (paper)` = dd_paper)

knitr::kable(dose_tbl,
             caption = "Correction factors and recommended dose reductions, model versus Mou 2024 supplemental Table S3.")
```

# Assumptions and deviations

The paper does not report several things this reduction needs, and it contains
a number of internal inconsistencies. Both are recorded here.

## Parameters not reported by the source

* **All distribution parameters and the absorption rate constant.**
  `lka`, `lvc`, `lvp`, `lq`, `lvc_dehydro`, `lvp_dehydro` and `lq_dehydro` are
  **not** in Mou 2024. They were recovered by digitizing the predicted mean
  curves of Figure 2(a) and fitting this reduction to them with every clearance
  fixed at its published value. A one-compartment parent could not fit the
  curve (the absorption rate ran to infinity); a two-compartment parent with a
  two-compartment metabolite was required.
* **Bioavailability.** `lfdepot` = 0.974 is derived, not published. It is the
  value that makes Table 1's clearance consistent with the reported single-dose
  AUC at the 70 kg reference weight.
* **Reference weight.** Table 1 gives clearance per kg but no reference
  individual. Supplemental Table S1's virtual-population median is 78.0 kg, but
  at 78 kg the paper's own clearance and AUC imply a bioavailability of 1.08,
  which is impossible; a bioavailability of at most 1 requires a reference
  weight of about 72 kg or less. The rounded standard of 70 kg is used.
* **Volume-weight scaling.** Volumes are scaled linearly with body weight. The
  paper gives no basis for this; linear scaling was chosen so that the
  typical-value half-life is weight-invariant, consistent with the per-kg
  clearance parameterization.
* **Residual error.** Not reported; `propSd` and `propSd_dehydro` are 0.10
  placeholders following the in-repo PBPK convention.
* **Between-subject variability.** None. The confidence bands in the paper's
  figures come from PK-Sim virtual-population sampling over anatomy and enzyme
  abundance, which is not a population-PK omega and is not transferable.
* **The risperidone inhibition constant.** Mou 2024 state that the DDI model
  was built "using inhibition constant (Ki)" but never give its value, and the
  risperidone compound parameters are cited to an external reference that is
  not part of this article. Risperidone is therefore not modelled as a
  perpetrator; its effect is encoded as the flat clearance multiplier the paper
  reports (a 0.9% AUC increase, i.e. no meaningful interaction, which is the
  paper's own conclusion). The published effect was dose-independent over
  risperidone 2-6 mg, so a binary indicator is sufficient.

## Internal inconsistencies in the source

* **Two CYP2D6 poor-metabolizer parameterizations.** The Results text and
  abstract report a PM/NM AUC fold change of 1.75, which corresponds to zeroing
  CYP2D6 and leaving CYP3A4 unchanged; supplemental Table S3 and Table 5 report
  1.371, which corresponds to Table 2's footnote-a CYP3A4 up-adjustment. Both
  are reproduced above. The packaged model uses Table 2 as printed because that
  is the basis of the paper's dose recommendations.
* **The poor-metabolizer metabolite curve cannot be reproduced from Table 2.**
  Figure 2(c) plots dehydro-aripiprazole plateauing near 3 ng/mL in poor
  metabolizers, roughly half the normal-metabolizer level. Table 2 implies the
  opposite: in a poor metabolizer all metabolism runs through CYP3A4, whose
  dehydrogenation share is higher, so the fraction of the dose converted to
  dehydro-aripiprazole rises from 0.40 to 0.55 and the metabolite exposure
  should rise by about 1.37-fold. No combination of the paper's published
  numbers produces the plotted curve. Note that the paper's own observed
  dehydro-aripiprazole data in Figure 2(c) lie well **above** its predicted
  curve, and the Results section acknowledges that dehydro-aripiprazole
  prediction in poor metabolizers is poor. The packaged model follows Table 2.
* **Three different Child-Pugh A effect sizes.** Supplemental Table S3 gives a
  DDZI AUC ratio of 1.078 (+8%); the Results text states that steady-state
  concentrations in mild hepatic impairment were "1.65 +/- 0.12 times" higher
  while simultaneously concluding "there is no difference in the drug exposure
  between the two populations"; and Figure 4's reported AUC of 5485.65 ng*h/mL
  after 15 mg implies about +26% relative to the paper's own 20 mg AUC scaled
  to 15 mg. The Table S3 value is encoded because it is the one that produces
  the correction factor of 0.93 the paper recommends. As a consequence the
  15 mg Child-Pugh A simulation under-predicts Figure 4's aripiprazole AUC by
  about 15% and over-predicts its dehydro-aripiprazole AUC by about 25%.
* **Table 3 physiology does not compose to the reported effect.** Table 3's
  Child-Pugh A factors (fractional liver mass 0.69, CYP2D6 abundance 0.76,
  CYP3A4 abundance 0.79, unbound fraction 1.23) would, under a well-stirred
  low-extraction reduction, raise AUC by roughly 50%, not the 8% the paper
  reports. The empirical Table S3 ratio is used instead of composing Table 3.
* **The DDI and DDZI effects are not multiplicative.** Table 5's combined
  Child-Pugh A plus risperidone AUC ratio is 1.126, whereas the individual
  effects multiply to 1.088. Mou 2024 describe this explicitly as a synergistic
  effect. The packaged model applies the two covariate effects
  multiplicatively, so it reproduces each effect alone exactly but
  under-predicts the combination by about 3%.
* **Table 4 versus Figure 3.** The Results text states that "the MPE values of
  the two models of CYP2D6 PM and CYP2D6 UM were greater than 10% when
  predicting dehydro-aripiprazole", but Table 4 reports PM and UM
  dehydro-aripiprazole MPE as -1.51% and -8.41%, while the largest value in the
  column is IM at -21.92%. This does not affect the model.

## Scope

The model covers aripiprazole and dehydro-aripiprazole disposition with CYP2D6
phenotype, Child-Pugh A hepatic impairment, and risperidone coadministration.
It does not reimplement the PK-Sim whole-body PBPK structure, the Weibull
dissolution model (b = 0.91, T50 = 5 min), or the risperidone compound model.
