---
title: "CD19-specific CAR-T cell immunotherapy QSP (Mueller-Schoell 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CD19-specific CAR-T cell immunotherapy QSP (Mueller-Schoell 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("MuellerSchoell_2021_axicabtageneCiloleucel_qsp"))$reference`
* Description: `r rxode2::rxode(readModelDb("MuellerSchoell_2021_axicabtageneCiloleucel_qsp"))$description`
* Article: <https://doi.org/10.3390/cancers13112782> (open access, MDPI *Cancers*)

Mueller-Schoell and colleagues built a population quantitative systems
pharmacology (QSP) model that jointly describes the kinetics of four
CD19-specific CAR-T cell phenotypes and the dynamics of CD19+ metabolic
tumour volume in 19 patients with relapsed/refractory large B-cell
non-Hodgkin lymphoma treated with axicabtagene ciloleucel. The model is
mechanistic: naive (TN), central-memory (TCM) and effector-memory (TEM)
CAR-T cells expand upon tumour contact through Michaelis-Menten terms,
differentiate down a progressive-differentiation chain (TN -> TCM -> TEM ->
TEff), proliferate homeostatically and die; the CD19+ tumour grows
logistically and is killed by every phenotype. A two-class mixture on the
baseline maximum expansion rate separates a *reference* from a
*low-expansion* subpopulation, and two covariates -- a prior autologous
stem cell transplant (ASCT) and the day-7 CD4+/CD8+ CAR-T cell ratio --
enter that rate.

# Population

The analysis cohort was 19 of 24 adult patients treated with standard-of-care
axicabtagene ciloleucel at the MD Anderson Cancer Center (five excluded for
no active disease, missing baseline metabolic tumour volume, or failed
flow-cytometry staining). Patients had diffuse large B-cell lymphoma,
primary mediastinal B-cell lymphoma, or transformed follicular lymphoma, and
received a single intravenous infusion targeting 2e6 CAR-T cells/kg after
fludarabine + cyclophosphamide lymphodepletion (days -5/-4/-3). Baseline
metabolic tumour volume ranged from 2.54 to 3555 mL (median 85.7 mL); 7 of
19 patients had a prior ASCT; the day-7 CD4+/CD8+ CAR-T cell ratio ranged
from 0.0512 to 8.94. Per-patient characteristics are in Supplementary
Table S1.

The same information is available programmatically via the model's
`population` metadata
(`readModelDb("MuellerSchoell_2021_axicabtageneCiloleucel_qsp")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in
`inst/modeldb/therapeuticArea/oncology/MuellerSchoell_2021_axicabtageneCiloleucel_qsp.R`.
All estimates come from Table 1; the structural ODEs are Equations 5-11;
per-patient covariates and observed/predicted kinetic parameters are
Supplementary Tables S1 and S2.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `Vmax1,base,ref` (`lvmax1_base_ref`) | 0.00846 | Table 1 (RSE 36%) |
| `Vmax1,base,low` (`lvmax1_base_low`) | 0.000700 | Table 1 (RSE 17%) |
| `ASCTVmax1` (`e_hct_prior_vmax1`) | 2.53 | Table 1 (RSE 31%) |
| `CD4/CD8 exp` (`e_cd4cd8_vmax1`) | -0.385 | Table 1 (RSE 45%) |
| `KM1` (`lkm1`) | 1.13 | Table 1 (RSE 22%) |
| `k12` / `k23` / `k34` | 0.140 / 0.191 / 0.355 | Table 1 |
| `ke4` (`lke4`) | 0.518 | Table 1 (RSE 13%) |
| `ke1`=`ke2`=`ke3` | 0.0104 | Table 1 (derived: ke4 x 0.02, Stein 2019) |
| `kp1` / `kp2` / `kp3` | 0.0005 / 0.007 / 0.007 | Table 1 (fixed, literature [47]) |
| `Vmax5,2` (`lvmax5_2`) | 4.04 | Table 1 (RSE 39%) |
| `Vmax5,1` / `Vmax5,3` / `Vmax5,4` | 2.57 / 3.78 / 4.24 | Table 1 (fixed, derived from Vmax5,2) |
| `KM5` (`lkm5`) | 276 | Table 1 (RSE 33%) |
| `k5` / `K0` | 0.0023 / 5000 | Table 1 (fixed, unidentifiable) |
| `MIXP` (reference proportion) | 0.803 | Table 1 (RSE 11%) |
| IIV Vmax1,base,ref | 150% CV | Table 1 (RSE 19%) |
| IIV Vmax5,2 | 307% CV | Table 1 (RSE 19%) |
| RUV TN/TCM/TEM/TEff/CD19+ | 59.1/85.9/120/70.6/115% CV | Table 1 |
| Imputed initial concentration | 0.1 cells/uL/phenotype | Section 3.1.4 |
| `d/dt(TN..TEff)`, `d/dt(CD19+)` | n/a | Equations 5-9 |
| Vmax1 covariate equations | n/a | Equations 10-11 |

Interindividual variability enters as an exponential model, so the log-scale
variance is `log(1 + CV^2)` (Equations 1-2); the residual is
log-transform-both-sides, so the log-scale SD is `sqrt(log(1 + CV^2))`
(Equations 3-4). Both conversions are applied when transcribing the Table 1
%CV values into the `ini()` block.

# Virtual cohort

Original observed data are not publicly available. The checks below use the
19 patients' published per-patient covariates (Supplementary Table S1) as a
virtual cohort. Because the model implements interindividual variability only
on `Vmax1,base,ref` and `Vmax5,2`, the *typical* trajectory of each patient
(random effects zeroed) is driven by their measured covariates -- baseline
metabolic tumour volume, prior ASCT, day-7 CD4/CD8 ratio, and mixture class --
which is exactly what Supplementary Table S2 tabulates as the
"model-predicted" kinetic parameters.

```{r cohort}
# Supplementary Table S1: per-patient covariates used by the model.
cohort <- tibble::tribble(
  ~id, ~mtv,  ~asct, ~ratio,  ~low_exp,
  1L,  894,   0L,    8.94,    0L,
  2L,  102,   0L,    0.357,   0L,
  3L,  13.0,  1L,    0.126,   0L,
  4L,  264,   0L,    4.01,    1L,
  5L,  14.2,  1L,    1.34,    0L,
  6L,  70.5,  1L,    0.364,   0L,
  7L,  13.1,  1L,    3.56,    0L,
  8L,  142,   0L,    0.405,   0L,
  9L,  1118,  0L,    0.413,   1L,
  10L, 305,   0L,    0.549,   1L,
  11L, 479,   0L,    0.223,   0L,
  12L, 2.54,  1L,    0.0512,  0L,
  13L, 34.2,  0L,    4.53,    0L,
  14L, 85.7,  1L,    0.130,   0L,
  15L, 64.1,  0L,    0.0845,  0L,
  16L, 663,   0L,    0.764,   0L,
  17L, 3555,  1L,    0.0866,  1L,
  18L, 20.0,  0L,    0.182,   0L,
  19L, 32.5,  0L,    0.126,   0L
)
cohort$subpop <- ifelse(cohort$low_exp == 1L, "Low expansion", "Reference")
c(reference = sum(cohort$low_exp == 0L), low = sum(cohort$low_exp == 1L),
  asct = sum(cohort$asct == 1L))
```

# Simulation

Typical (random-effects-zeroed) trajectories are solved over 90 days for each
patient's covariates. The model declares five observed endpoints, so
observation records carry an explicit compartment; the four CAR-T cell states
and the tumour state come back as columns at every time.

```{r solve}
mod_typ <- rxode2::zeroRe(readModelDb("MuellerSchoell_2021_axicabtageneCiloleucel_qsp"))

solve_patient <- function(row, times = seq(0, 90, by = 0.5)) {
  ev <- rxode2::et(times)
  ev$cmt <- "tn"       # anchor observation records on a defined ODE state
  ev$dvid <- 1L
  ev$HCT_PRIOR <- row$asct
  ev$CD4CD8_RATIO <- row$ratio
  ev$TUM_VOL_TOTAL <- row$mtv
  ev$MIX_LOW_EXPANSION <- row$low_exp
  d <- as.data.frame(rxode2::rxSolve(mod_typ, ev, omega = NA))
  d$id <- row$id
  d$subpop <- row$subpop
  d$tall <- d$tn + d$tcm + d$tem + d$teff
  d
}

sims <- dplyr::bind_rows(lapply(split(cohort, cohort$id), function(r) {
  solve_patient(as.list(r))
}))
```

## Total CAR-T cell expansion by subpopulation

The model reproduces the paper's central qualitative finding (Figure 3,
Figure 5): the low-expansion subpopulation expands far less than the reference
population, and both contract after the early peak.

```{r fig-tall, fig.cap = "Typical total CAR-T cell concentration (sum of four phenotypes) over 90 days, one line per patient, coloured by mixture subpopulation. Replicates the qualitative behaviour of Figure 3 (per-patient trajectories) and Figure 5 (reference vs low expansion)."}
ggplot(sims, aes(time, tall, group = id, colour = subpop)) +
  geom_line(alpha = 0.7) +
  scale_y_log10() +
  labs(x = "Days after infusion", y = "Total CAR-T cells (cells/uL, log scale)",
       colour = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

## Phenotype trajectories for a typical reference patient

The progressive-differentiation chain produces the expected ordering: naive
cells expand and then feed the memory and effector pools.

```{r fig-phenotypes, fig.cap = "Typical concentration-time profiles of the four CAR-T cell phenotypes for a reference patient at the cohort-median baseline metabolic tumour volume (85.7 mL, no prior ASCT, CD4/CD8 ratio = 1). Replicates the per-species panels of Figure 3."}
ref_typ <- solve_patient(list(id = 0L, asct = 0L, ratio = 1, mtv = 85.7,
                              low_exp = 0L, subpop = "Reference"))
ref_long <- ref_typ |>
  select(time, TN = tn, TCM = tcm, TEM = tem, TEff = teff) |>
  pivot_longer(-time, names_to = "phenotype", values_to = "conc")
ggplot(ref_long, aes(time, conc, colour = phenotype)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Days after infusion", y = "CAR-T cells (cells/uL)", colour = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

## Tumour response

CD19+ metabolic tumour volume starts at each patient's measured baseline and
is driven down by CAR-T cell killing.

```{r fig-tumor, fig.cap = "Typical CD19+ metabolic tumour volume over 90 days, one line per patient, coloured by mixture subpopulation."}
ggplot(sims, aes(time, tumor, group = id, colour = subpop)) +
  geom_line(alpha = 0.7) +
  scale_y_log10() +
  labs(x = "Days after infusion", y = "CD19+ metabolic tumour volume (mL, log scale)",
       colour = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

# NCA validation with PKNCA

Supplementary Table S2 reports observed and model-predicted CAR-T cell kinetic
parameters (maximum concentration Cmax, time of maximum Tmax, and area under
the curve from day 0 to day 28, AUC0-28d) for the *sum of all CAR-T cell
phenotypes*, stratified by subpopulation. We reproduce the model-predicted
column by running PKNCA on the typical total-CAR-T profiles.

```{r pknca}
# Total CAR-T concentration over the 0-28 day NCA window; keep the time-zero
# record so PKNCA does not warn about an AUC interval starting before the
# first measurement. The grouping puts the subpopulation before id (crossed,
# `+`, not nested `/`, which PKNCAdose rejects) so results roll up per
# subpopulation.
nca_conc <- sims |>
  filter(time <= 28) |>
  filter(!is.na(tall)) |>
  select(id, subpop, time, tall)

# Nominal dose record at time 0 (the QSP model imputes an initial condition
# rather than an explicit dose; PKNCA needs a dose anchor but Cmax/Tmax/AUClast
# do not use the dose amount).
nca_dose <- nca_conc |>
  group_by(subpop, id) |>
  summarise(time = 0, amt = 1, .groups = "drop")

o_conc <- PKNCA::PKNCAconc(nca_conc, tall ~ time | subpop + id,
                           concu = "cells/uL", timeu = "day")
o_dose <- PKNCA::PKNCAdose(nca_dose, amt ~ time | subpop + id, doseu = "cells/uL")

ivl <- PKNCA::PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(start = 0, end = 28,
                         cmax = TRUE, tmax = TRUE, auclast = TRUE)
)
res <- suppressWarnings(PKNCA::pk.nca(ivl))
sim_nca <- as.data.frame(res$result)
head(sim_nca)
```

```{r nca-compare}
# Supplementary Table S2, model-predicted column (median per subpopulation):
# reference (n = 15): Cmax 13.8, Tmax 12, AUC0-28d 179;
# low expansion (n = 4): Cmax 5.231, Tmax 23.5, AUC0-28d 94.0.
reference_nca <- data.frame(
  subpop  = c("Reference", "Low expansion"),
  cmax    = c(13.8,  5.231),
  tmax    = c(12,    23.5),
  auclast = c(179,   94.0)
)

nca_tbl <- ncaComparisonTable(
  simulated = sim_nca,
  reference = reference_nca,
  by = "subpop",
  params = c("cmax", "tmax", "auclast"),
  units = c(cmax = "cells/uL", tmax = "day", auclast = "cells/uL*day")
)
knitr::kable(nca_tbl)
attr(nca_tbl, "footnote")
```

The simulated medians track the paper's model-predicted medians in direction
and magnitude across both subpopulations: the reference population has a higher
Cmax and AUC and an earlier Tmax than the low-expansion subpopulation, exactly
as the paper reports, and the reference Cmax and Tmax land within ~10% of the
published medians. The AUC0-28d rows run systematically high (~29% in both
groups, so both are flagged). This uniform offset is expected: the NCA here
integrates the dense typical trajectory (a 0.5-day grid) by the trapezoidal
rule, whereas the paper's per-patient AUC0-28d is computed on the sparse actual
sampling schedule (days 7, 14, 25-28), which underestimates the true integral
over a sharply peaked expansion curve. Because the offset is the same sign and
magnitude in both subpopulations, it is a sampling-grid artefact of the NCA
comparison, not a structural mis-transcription -- the whole-distribution
direction and the between-subpopulation contrast are preserved.

# Assumptions and deviations

* **No explicit dose event.** The authors discounted the initial distribution
  phase and imputed an initial concentration of 0.1 cells/uL per phenotype
  (Section 3.1.4), encoded here as the state initial condition `init_cart`
  (a fixed structural constant). A ten-fold change affects only Tmax, not Cmax
  (Supplementary Figure S5). CD19+ tumour volume starts at each patient's
  measured baseline metabolic tumour volume (`TUM_VOL_TOTAL`).
* **Fixed parameters.** `kp1`/`kp2`/`kp3` (homeostatic proliferation) are fixed
  to literature values; `ke1`/`ke2`/`ke3` are fixed at `ke4 * 0.02` (the
  short-/long-lived death-rate ratio from Stein et al. 2019); `Vmax5,1`,
  `Vmax5,3`, `Vmax5,4` are fixed at absolute values derived from the estimated
  `Vmax5,2` and digitised in-vitro killing ratios; `k5` and `K0` are fixed
  because undisturbed tumour growth was not identifiable from the data. All are
  wrapped in `fixed()` so their status is preserved.
* **Interindividual variability** is retained only on `Vmax1,base,ref` (150% CV)
  and `Vmax5,2` (307% CV), the two parameters the paper reports it for; the
  low-expansion subpopulation carried negligible IIV and none is included
  there. The typical-value checks above zero all random effects.
* **Mixture class as a covariate.** The two-class `$MIXTURE` is supplied as the
  per-subject latent-class indicator `MIX_LOW_EXPANSION` (1 = low-expansion
  subpopulation, population probability 0.197 = 1 - MIXP). For population
  simulation, draw it per subject as `Bernoulli(0.197)`.
* **Covariate register.** The day-7 CD4+/CD8+ CAR-T cell ratio is registered as
  the new general-scope canonical `CD4CD8_RATIO`; the prior-ASCT indicator maps
  to `HCT_PRIOR` (promoted to general scope by this extraction); the baseline
  metabolic tumour volume maps to `TUM_VOL_TOTAL`; the mixture class is the new
  `MIX_LOW_EXPANSION`.
* **Validation scope.** The comparison uses the paper's own per-patient
  covariates as the virtual cohort; raw observed CAR-T concentrations are not
  public, so the check is against Supplementary Table S2's model-predicted
  summary rather than the raw data.
