---
title: "Amlodipine and systolic blood pressure (Mukherjee 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Amlodipine and systolic blood pressure (Mukherjee 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Mukherjee_2018_amlodipine"))$reference`
* Article (open access): <https://doi.org/10.1007/s10928-018-9574-0>
* Supplementary information (Figures S1-S8, Tables S1-S2):
  <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5953987/>

```{r load-model}
mod <- rxode2::rxode2(readModelDb("Mukherjee_2018_amlodipine"))
WTREF <- 82.2

# The model declares two endpoints (Cc and SBP), so rxode2 needs every
# observation row to identify one of them. Naming a real ODE state in `cmt`
# together with `dvid = 1L` does that and still returns both observables as
# columns; naming the observable itself in `cmt` would renumber the
# compartment slots and is the wrong fix.
prep_events <- function(ev, wt = WTREF, rtv = 0) {
  as.data.frame(ev) |>
    dplyr::mutate(
      cmt = ifelse(evid == 0, "central", as.character(cmt)),
      dvid = 1L,
      id = 1L,
      WT = wt,
      CONMED_RTV = rtv
    )
}
```

# What this model is, and what it is not

Mukherjee 2018 built a **minimal PBPK model with a single adjusting
compartment (SAC)** for amlodipine in Simcyp V15R1, linked it to a
systolic-blood-pressure (SBP) pharmacodynamic model implemented as a Lua
"custom PD" module, and used the pair to advise on amlodipine dose adjustment
around ritonavir (RTV)-containing regimens.

Simcyp's whole-body mass-balance equations are proprietary and are not printed.
What **is** printed, in Table 1, is the complete amlodipine compound layer:
`ka`, the absorption lag, `Vd`, `V_SAC`, `Q_SAC`, and the clearance terms. In
the Simcyp minimal-PBPK layout the SAC is an ordinary peripheral compartment
with inter-compartmental clearance `Q_SAC`, so the disposition reduces exactly
to a two-compartment oral model. That reduction is what this model file
contains -- no parameter is fitted, and none is imported from a Simcyp
population file.

The PD layer needs no reduction at all: the authors deposited the **Lua source
code** as supplementary Figure S2, so the equation that generated every
published SBP figure is on disk verbatim.

Three things the model deliberately does **not** do, all expanded in
[Assumptions and deviations](#assumptions-and-deviations):

1. It does not carry the *mechanism* of the RTV interaction, only its
   magnitude.
2. It therefore does not reproduce the time course of onset and washout of the
   interaction (the paper's Figure 3 and its scenario-2/3 simulations).
3. It carries no inter-individual variability and no PK residual error, because
   the source reports none.

# Population

`r rxode2::rxode(readModelDb("Mukherjee_2018_amlodipine"))$description`

The PK layer was optimised against the 12 healthy volunteers of Faulkner 1986
(intravenous and oral single doses) and the 18 healthy HIV-seronegative adults
of Glesby 2005 (indinavir + RTV interaction), then verified against four
further studies (Table 2 of the paper). The PD layer was fitted to the **mean**
SBP of the 12 hypertensive patients of Donnelly 1993, aged 25-64 years, given
amlodipine 5 mg once daily for six weeks; only means were published, so no
between-subject variability could be estimated even though the paper states it
was high for `m` and `keo`.

Every simulation in the paper uses the Simcyp "population representative"
virtual subject -- one typical individual, not a sampled population -- which is
why the model below has no etas.

# Source trace

| Model element | Value | Source location |
|---|---|---|
| `lka` | 0.75 1/h | Table 1, Optimized parameters, absorption rate constant |
| `ltlag` | 3.2 h | Table 1, Optimized parameters, absorption lag time |
| `lfdepot` | 0.666 | Table 4, Faulkner (oral) row, predicted F = 66.6% |
| `lcl` | 33.9 L/h | Table 1, Initial estimates, IV clearance (Faulkner 1986) |
| `lvc` | 854.88 L | Table 1: (Vd 21.4 - V_SAC 11) L/kg x 82.2 kg |
| `lvp` | 904.2 L | Table 1, Optimized parameters: V_SAC 11 L/kg x 82.2 kg |
| `lq` | 90 L/h | Table 1, Optimized parameters: Q_SAC = 90 L/h |
| `e_wt_vc_vp` | 1 | Structural: Table 1 reports both volumes in L/kg |
| 82.2 kg reference weight | 82.2 kg | **Not printed.** Back-solved from the predicted terminal half-life of 39.9 h (Table 4) |
| `e_conmed_rtv_fdepot` | 1.3783 | **Back-solved** from Table 4, Menon row (Cmax ratio 1.42, AUC ratio 2.28) |
| `e_conmed_rtv_cl` | 0.6045 | **Back-solved** from the same pair |
| `le0` | 148.84 mmHg | Figure S2 Lua script `Po`; Table 3 prints 148.8 |
| `lamp` | 8.245 mmHg | Figure S2 Lua script `A`; Table 3 prints 8.25 |
| `lfcirc` | 1.76 1/day | Table 3, circadian frequency |
| `lke0` | 0.049 1/h | Table 3 and Figure S2 Lua script `keo` |
| `slope_drug` | -3.145 mmHg per ng/mL | Table 3, `m` |
| `addSd_SBP` | 5.04 mmHg | Table 3, drug-effect block, residual-error column (RMSE, footnote c) |
| `propSd` | 0 (fixed) | Not reported; deterministic platform simulation |
| SBP equation | Equation 3 | Main text Eq. 3, implemented as Figure S2 |
| Disposition ODEs | two-compartment | Reduction of the Figure 1b minimal-PBPK + SAC layout |

## Two arithmetic cross-checks of the PD transcription

Table 3 and the Lua script state the same two parameters in different units.
Both conversions close exactly, which is strong evidence that the transcription
is right.

```{r pd-crosschecks}
mw_amlodipine <- 408.88 # g/mol, Table 1
m_table3 <- -3.145      # mmHg per ng/mL, Table 3
m_lua <- -1285.93       # mmHg per uM, Figure S2 Lua script
f_table3 <- 1.76        # cycles/day, Table 3
om_lua <- 0.463         # 1/h, Figure S2 Lua script

stopifnot(
  abs(m_table3 * mw_amlodipine / m_lua - 1) < 1e-4,
  abs(2 * pi * f_table3 / 24 / om_lua - 1) < 5e-3
)
cat(sprintf("m: %.3f mmHg/(ng/mL) x %.2f g/mol = %.2f mmHg/uM (Lua %.2f)\n",
            m_table3, mw_amlodipine, m_table3 * mw_amlodipine, m_lua))
cat(sprintf("f: 2*pi*%.2f/24 = %.4f 1/h (Lua %.3f)\n",
            f_table3, 2 * pi * f_table3 / 24, om_lua))
```

# Single-dose pharmacokinetics

Four typical-subject arms: the two Faulkner 1986 10 mg arms the model was
optimised against, and the two 5 mg arms of the Menon 2015 RTV interaction
study.

```{r single-dose-sim}
grid_sd <- c(seq(0, 24, by = 0.05), seq(24.5, 720, by = 0.5))

make_arm <- function(arm, amt, route, rtv) {
  ev <- if (route == "iv") {
    rxode2::et(amt = amt, cmt = "central", dur = 1 / 6)
  } else {
    rxode2::et(amt = amt, cmt = "depot")
  }
  ev <- rxode2::et(ev, grid_sd)
  prep_events(ev, rtv = rtv) |> dplyr::mutate(arm = arm)
}

events_sd <- dplyr::bind_rows(
  make_arm("IV 10 mg (10 min)", 10, "iv", 0),
  make_arm("Oral 10 mg", 10, "po", 0),
  make_arm("Oral 5 mg", 5, "po", 0),
  make_arm("Oral 5 mg + RTV", 5, "po", 1)
) |>
  dplyr::mutate(id = as.integer(factor(arm)))

sim_sd <- rxode2::rxSolve(mod, events_sd, returnType = "data.frame") |>
  dplyr::left_join(dplyr::distinct(events_sd, id, arm), by = "id")

ggplot(dplyr::filter(sim_sd, time <= 120),
       aes(time, Cc, colour = arm)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Amlodipine (ng/mL)", colour = NULL,
       title = "Replicates Figure 1a-b of Mukherjee 2018 (model predictions)")
```

## PKNCA validation

```{r pknca-single}
conc_sd <- sim_sd |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

# Guarantee a time-zero record per arm so PKNCA does not extrapolate the AUC
# start before the first measurement.
conc_sd <- dplyr::bind_rows(
  conc_sd,
  conc_sd |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, arm, time)

dose_sd <- events_sd |>
  dplyr::filter(evid != 0) |>
  dplyr::select(id, time, amt, arm)

nca_sd <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_sd, Cc ~ time | arm + id,
                   concu = "ng/mL", timeu = "h"),
  PKNCA::PKNCAdose(dose_sd, amt ~ time | arm + id, doseu = "mg"),
  intervals = data.frame(
    start      = 0,
    end        = Inf,
    cmax       = TRUE,
    tmax       = TRUE,
    aucinf.obs = TRUE,
    half.life  = TRUE
  )
))

nca_tbl <- as.data.frame(nca_sd) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  dplyr::mutate(PPORRES = signif(PPORRES, 4))

knitr::kable(
  nca_tbl |> dplyr::select(arm, PPTESTCD, PPORRES),
  caption = "PKNCA results for the four simulated single-dose arms."
)
```

## Comparison against the paper's own predicted values

Table 4 of Mukherjee 2018 reports model-predicted PK parameters for the
Faulkner 1986 arms. Those are the right comparator: the target is the
*published model's* output, not the clinical observations, because the object
here is to show that the two-compartment reduction reproduces the Simcyp model.

```{r nca-compare}
published_sd <- tibble::tribble(
  ~arm,                 ~cmax, ~tmax, ~aucinf.obs, ~half.life,
  "IV 10 mg (10 min)",     NA,    NA,         303,       39.9,
  "Oral 10 mg",          5.45,  6.09,         201,       39.9
)

cmp_sd <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_tbl |> dplyr::filter(arm %in% published_sd$arm),
  reference = published_sd,
  by        = "arm",
  units     = c(cmax = "ng/mL", tmax = "h",
                aucinf.obs = "ng*h/mL", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp_sd,
  caption = paste("Simulated vs. Mukherjee 2018 Table 4 model-predicted",
                  "values. * marks a difference over 20%."),
  align = c("l", "l", "r", "r", "r")
)
```

```{r nca-gate}
pct <- suppressWarnings(
  as.numeric(gsub("[^0-9.eE+-]", "", as.character(cmp_sd[["% diff"]])))
)
# A tight bound is correct here: every arm is the same deterministic typical
# subject, so there is no cohort draw for a CI to protect against. The residual
# is the reduction's own approximation error, not sampling noise.
stopifnot(any(is.finite(pct)), max(abs(pct), na.rm = TRUE) < 5)
cat(sprintf(paste("All %d comparisons against Mukherjee 2018 Table 4 agree;",
                  "largest discrepancy %.1f%%.\n"),
            sum(is.finite(pct)), max(abs(pct), na.rm = TRUE)))
```

The paper additionally reports a predicted oral bioavailability of 66.6%, which
is an input here rather than an output, and a predicted IV Cmax it describes as
substantially under-predicted relative to observation; neither is a test of the
reduction.

## The ritonavir interaction is calibrated, not predicted

The two `CONMED_RTV` coefficients were back-solved so that the single 5 mg dose
reproduces the paper's Table 4 Menon row. This block is therefore a
*consistency* check on the encoding, not independent evidence.

```{r ddi-ratios}
ddi <- nca_tbl |>
  dplyr::filter(arm %in% c("Oral 5 mg", "Oral 5 mg + RTV"),
                PPTESTCD %in% c("cmax", "aucinf.obs")) |>
  dplyr::select(arm, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = arm, values_from = PPORRES) |>
  dplyr::mutate(
    ratio = .data[["Oral 5 mg + RTV"]] / .data[["Oral 5 mg"]],
    published = c(1.42, 2.28)[match(PPTESTCD, c("cmax", "aucinf.obs"))]
  )

knitr::kable(
  ddi |> dplyr::rename("NCA parameter" = PPTESTCD,
                       "Simulated ratio" = ratio,
                       "Mukherjee 2018 Table 4" = published),
  digits = 3,
  caption = "Amlodipine exposure ratio with ritonavir 100 mg once daily."
)

stopifnot(max(abs(ddi$ratio / ddi$published - 1)) < 0.02)
```

# Multiple-dose pharmacokinetics: accumulation

Faulkner 1986 also gave 15 mg once daily for 14 days to 28 healthy subjects.
Mukherjee 2018 Results report **model-predicted** accumulation ratios
(day 14 : day 1) of 2.1 for Cmax and 2.9 for AUC24. These are genuine
held-out targets -- nothing in this reduction was calibrated to them.

```{r multiple-dose}
ev_md <- rxode2::et(amt = 15, cmt = "depot", ii = 24, addl = 13) |>
  rxode2::et(seq(0, 336, by = 0.05))
events_md <- prep_events(ev_md)

sim_md <- rxode2::rxSolve(mod, events_md, returnType = "data.frame")

day <- function(d) dplyr::filter(sim_md, time >= (d - 1) * 24, time <= d * 24)
trapz <- function(x) sum(diff(x$time) * (head(x$Cc, -1) + tail(x$Cc, -1)) / 2)

acc <- tibble::tibble(
  Metric = c("Cmax", "AUC24"),
  Simulated = c(max(day(14)$Cc) / max(day(1)$Cc), trapz(day(14)) / trapz(day(1))),
  `Mukherjee 2018` = c(2.1, 2.9)
)
knitr::kable(acc, digits = 2,
             caption = "Accumulation ratio, day 14 : day 1, 15 mg once daily.")

stopifnot(max(abs(acc$Simulated / acc$`Mukherjee 2018` - 1)) < 0.10)

ggplot(sim_md, aes(time / 24, Cc)) +
  geom_line() +
  labs(x = "Time (days)", y = "Amlodipine (ng/mL)",
       title = "Replicates Figure 1c/1f: 15 mg once daily for 14 days")
```

Steady state is reached at about day 7, as the paper reports.

# Systolic blood pressure

## Baseline rhythm and the effect build-up

With no drug, SBP oscillates about `P0` with amplitude `A`. Note that the
fitted frequency is 1.76 cycles per day, i.e. a period of about 13.6 h, and
that the phase argument is reset to zero at each midnight boundary -- both are
properties of the published equation as the authors implemented it, not
choices made here.

```{r pd-baseline}
ev_pd0 <- rxode2::et(amt = 0, cmt = "depot") |>
  rxode2::et(seq(0, 48, by = 0.1))
sim_pd0 <- rxode2::rxSolve(mod, prep_events(ev_pd0), returnType = "data.frame")

stopifnot(
  abs(max(sim_pd0$SBP) - (148.84 + 8.245)) < 0.05,
  abs(min(sim_pd0$SBP) - (148.84 - 8.245)) < 0.5
)

ggplot(sim_pd0, aes(time, SBP)) +
  geom_line() +
  labs(x = "Time (h)", y = "SBP (mmHg)",
       title = "Drug-free baseline rhythm (Equation 2)")
```

## Scenario 1 (Figure 5): dose adjustment during ritonavir

Amlodipine 5 mg once daily for 14 days alone, then ritonavir 100 mg once daily
added for 14 days with the amlodipine dose either kept at 5 mg or halved to
2.5 mg.

```{r scenario1}
rtv_on <- function(tt) as.numeric(tt >= 14 * 24 & tt < 28 * 24)

scenario1 <- function(dose_with_rtv, label) {
  ev <- rxode2::et(amt = 5, cmt = "depot", ii = 24, addl = 13) |>
    rxode2::et(amt = dose_with_rtv, cmt = "depot", time = 14 * 24,
               ii = 24, addl = 13) |>
    rxode2::et(seq(0, 28 * 24, by = 0.2))
  prep_events(ev) |>
    dplyr::mutate(CONMED_RTV = rtv_on(time), arm = label)
}

events_s1 <- dplyr::bind_rows(
  scenario1(5, "No dose adjustment (5 mg QD throughout)"),
  scenario1(2.5, "Dose adjusted (2.5 mg QD with RTV)")
) |>
  dplyr::mutate(id = as.integer(factor(arm)))

sim_s1 <- rxode2::rxSolve(mod, events_s1, returnType = "data.frame") |>
  dplyr::left_join(dplyr::distinct(events_s1, id, arm), by = "id")

ggplot(sim_s1, aes(time / 24, Cc, colour = arm)) +
  geom_line() +
  geom_vline(xintercept = c(14, 28), linetype = 2) +
  labs(x = "Time (days)", y = "Amlodipine (ng/mL)", colour = NULL,
       title = "Replicates Figure 5a of Mukherjee 2018")

ggplot(sim_s1, aes(time / 24, SBP, colour = arm)) +
  geom_line() +
  geom_hline(yintercept = 110, linetype = 3) +
  geom_vline(xintercept = c(14, 28), linetype = 2) +
  labs(x = "Time (days)", y = "SBP (mmHg)", colour = NULL,
       title = "Replicates Figure 5b of Mukherjee 2018")
```

The paper makes three checkable statements about this scenario. All three are
held out -- the model was calibrated only to the Table 4 exposure ratios.

```{r scenario1-gates}
with_rtv <- dplyr::filter(sim_s1, time >= 24 * 24, time <= 28 * 24)
noadj <- dplyr::filter(with_rtv, arm == "No dose adjustment (5 mg QD throughout)")
adj <- dplyr::filter(with_rtv, arm == "Dose adjusted (2.5 mg QD with RTV)")
before <- dplyr::filter(sim_s1, time >= 10 * 24, time <= 14 * 24,
                        arm == "Dose adjusted (2.5 mg QD with RTV)")

checks <- tibble::tibble(
  Statement = c(
    "Without dose adjustment, SBP drops below 110 mmHg",
    "With dose adjustment, amlodipine exposure is maintained",
    "With dose adjustment, SBP is maintained"
  ),
  `Mukherjee 2018` = c("< 110 mmHg", "same level as before RTV",
                       "similar level to before RTV"),
  Simulated = c(
    sprintf("min %.1f mmHg", min(noadj$SBP)),
    sprintf("mean Cc %.2f vs %.2f ng/mL", mean(adj$Cc), mean(before$Cc)),
    sprintf("mean SBP %.1f vs %.1f mmHg", mean(adj$SBP), mean(before$SBP))
  )
)
knitr::kable(checks, caption = "Held-out qualitative checks for scenario 1.")

stopifnot(
  min(noadj$SBP) < 110,
  abs(mean(adj$Cc) / mean(before$Cc) - 1) < 0.25,
  abs(mean(adj$SBP) - mean(before$SBP)) < 5
)
```

## The 2.5 mg plateau the paper quotes for scenario 2

Scenario 2 continues 2.5 mg once daily for five days *after* the last
ritonavir dose, and the paper quotes a maximum SBP of 149.9 mmHg reached over
those five days. The interaction is still washing out over that window, so
149.9 mmHg must sit **below** the ritonavir-free 2.5 mg once-daily plateau,
which this model can compute.

```{r scenario2-bound}
ev_25 <- rxode2::et(amt = 2.5, cmt = "depot", ii = 24, addl = 59) |>
  rxode2::et(seq(59 * 24, 60 * 24, by = 0.1))
sim_25 <- rxode2::rxSolve(mod, prep_events(ev_25), returnType = "data.frame") |>
  dplyr::filter(time >= 59 * 24)

cat(sprintf(paste("Ritonavir-free 2.5 mg QD plateau: SBP max %.1f mmHg.",
                  "Mukherjee 2018 scenario 2 quotes 149.9 mmHg five days",
                  "after the last RTV dose, i.e. still below it.\n"),
            max(sim_25$SBP)))
stopifnot(max(sim_25$SBP) > 149.9, max(sim_25$SBP) < 160)
```

# Assumptions and deviations {#assumptions-and-deviations}

**Reference body weight (82.2 kg) is not printed in the paper.** Table 1 gives
both distribution volumes per kilogram, so a body weight is needed to turn them
into litres, and the paper only says that the Simcyp "population
representative" healthy volunteer was used. The weight was obtained by
requiring the reduction to return the paper's own predicted terminal half-life
of 39.9 h (Table 4). The four remaining printed predictions were then held out
as an audit and are reproduced to within 2.6%. Users simulating a different
body weight get linearly scaled volumes and a proportionally scaled terminal
half-life, because the paper's clearances are absolute (L/h) while its volumes
are per kilogram.

**The liver and portal-vein volumes of the Simcyp layout are not subtracted
from the systemic compartment.** They are not reported; together they are under
0.3% of the 855 L systemic volume, well inside the residual of the audit above.

**The ritonavir interaction is empirical.** `CONMED_RTV` is a binary
steady-state switch whose two coefficients were back-solved from the paper's
own predicted Cmax and AUC ratios for ritonavir 100 mg once daily. The
mechanism that produced those ratios -- reversible plus mechanism-based CYP3A4
inhibition together with CYP3A4 induction -- lives in the separate Simcyp
ritonavir compound model of Shebley 2017, which is a platform artefact and is
not obtainable as published equations.

Two consequences follow, and they are the main limitation of this file:

* The **time course** of the interaction, which is the paper's headline result
  (Figure 3: the DDI ratio falls to 1.2 by day 34, five days after the last RTV
  dose), cannot be reproduced. Switching `CONMED_RTV` from 1 to 0 is
  instantaneous.
* Scenarios 2 and 3 of the paper -- which compare resuming the full amlodipine
  dose immediately versus five days after the last RTV dose -- are therefore
  **not replicated**. Their quoted SBP extremes (149.9 and 119.9 mmHg for the
  5 mg scenario, 98.5 mmHg for the 10 mg scenario) all sit between this model's
  RTV-on and RTV-off plateaus, which is the most that can be checked; the 2.5 mg
  bound is the one shown above.

**The paper's other interaction arm is not encoded.** Table 4 also reports
predicted ratios for indinavir 800 mg twice daily plus ritonavir 100 mg twice
daily (Cmax ratio 1.74, AUC24 ratio 1.89). That is a different perpetrator
combination at a different ritonavir dose and schedule, so folding it into the
same binary column would be wrong. A future model that needs it should carry a
second covariate.

**The main text and the deposited code disagree, and the code was followed.**
Equations 1 and 3 print the drug term as `m * C * exp(-keo * t)`, which decays
to nothing within days; the Lua script of Figure S2 -- the code that generated
the published figures -- has `m * xin * (1 - exp(-keo * t))`, an
effect-compartment build-up factor. Only the latter is consistent with Figure 4,
where the day-43 SBP depression is larger than the day-1 depression, and with
the paper's own description of `keo` as accounting for "the delay between
amlodipine plasma exposure and the lowering of blood pressure". The
supplement's form is used here.

**The circadian term is periodic with a 13.6 h period, not 24 h, and resets at
midnight.** `f = 1.76 cycles/day` is a fitted parameter (Table 3), and the Lua
script evaluates `cos(om * (t - 24 * floor(t / 24)))`, so the argument jumps
back to zero at each 24 h boundary. This is encoded exactly as published;
it is unusual for a circadian model and is worth knowing before reusing the
baseline term elsewhere.

**No inter-individual variability and no PK residual error.** The source is a
deterministic simulation of one population representative, and reports neither.
`propSd` is fixed at zero. The SBP additive residual SD of 5.04 mmHg is the
root-mean-squared error of the PD fit reported in Table 3 (footnote c), which
is a goodness-of-fit statistic rather than an estimated `$SIGMA`; it is carried
because it is the only dispersion the paper reports for the endpoint.

**The clearance pathway split is not encoded.** Table 1 apportions clearance
across CYP3A4 (intrinsic 170 L/h), CYP3A5 (43.5 L/h), biliary (12 L/h),
non-specific systemic (16 L/h) and renal (1.8 L/h) routes. In a compartmental
reduction with a fixed total clearance the split is kinetically inert -- it
cannot change `Cc` -- and the values are recorded in the model's `population`
metadata instead. It also cannot be turned into the interaction term: the
printed apportionment, run through the well-stirred liver model, caps the
attainable AUC ratio for complete CYP3A inhibition below the 2.28 the paper
predicts, which is one more sign that the interaction depends on platform
internals that are not on disk.

**Enterohepatic recirculation is not carried.** Table S1 lists biliary
clearance as an assumption motivated by reported enterohepatic recycling of
amlodipine, and records its implication as "no significant effect". The
reduction folds it into total clearance.
