---
title: "Mulyukov_2018_ranibizumab"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mulyukov_2018_ranibizumab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
```

```{r setup}
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Mulyukov_2018_ranibizumab"))$meta$reference`
* Description: `r rxode2::rxode2(readModelDb("Mulyukov_2018_ranibizumab"))$meta$description`
* Article: <https://doi.org/10.1002/psp4.12322>

Mulyukov 2018 develops an indirect-response nonlinear mixed-effects model of
best-corrected visual acuity (BCVA; ETDRS letters) under ranibizumab treatment
in anti-VEGF-naive neovascular age-related macular degeneration (nAMD). The
dynamics are governed by

$$\frac{dg_i(t)}{dt} = k_{\text{in},i}\left[1 + E_{\max,i}(t)\,\frac{C_i(t)}{EC_{50} + C_i(t)}\right] - k_{\text{out},i}\,g_i(t)$$

with $g_i(0) = g_{i,0}$ and a time-dependent maximum effect

$$E_{\max,i}(t) = E_{\max,i}^{ss} + \Delta E_{\max,i}^0\,e^{-k_{E_{\max}} t}.$$

The vitreous PK is not estimated in this paper. It is borrowed from a previous
population PK analysis (paper reference 20) and fixed at a first-order
elimination process with $k_{\text{elim}} = 0.077$/day ($t_{1/2} = 9$ days) and
a 4 mL vitreous volume. No IIV is carried on the PK parameters because
vitreous PK samples were not collected in the four development studies.

# Population

The model was developed on pooled individual data from four phase-III
ranibizumab clinical studies — ANCHOR, MARINA, PIER, and EXCITE — totalling
**1,524 treatment-naive patients** and **29,754 BCVA observations**. Inclusion
criteria common to all four studies: age ≥ 50 years with a study-eye BCVA of
25–70 ETDRS letters (nAMD of any lesion type across the pool). Baseline BCVA
across studies was 54 ± 13 letters (range 3–84); baseline age 77 ± 7.5 years
(range 52–96); ~40% male. Dosing regimens in the development dataset include
0.3 mg and 0.5 mg intravitreal ranibizumab, monthly (q4w) or quarterly (q12w)
after three monthly loading injections, along with sham comparator arms.
The PDT (verteporfin) control arm of ANCHOR and the year-2 PIER data were
excluded from the modelled dataset (see Mulyukov 2018 Methods — Clinical
studies and Table 1 for full study-level demographics). An external HARBOR
cohort (ranibizumab 0.5 mg and 2.0 mg q4w, summary data only) was used as a
predictive check and is **not** part of the fitted data.

The same information is available programmatically via the model's
`population` metadata:

```{r population}
str(rxode2::rxode2(readModelDb("Mulyukov_2018_ranibizumab"))$meta$population)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Mulyukov_2018_ranibizumab.R`. The table
below collects the equations and parameters in one place for reviewer audit.

| Equation / parameter               | Value                              | Source location                           |
|------------------------------------|------------------------------------|-------------------------------------------|
| PK ODE `d/dt(central) = -kel·central` | n/a                             | Methods — Model development, paragraph above Eq. 1 (first-order vitreous elimination, ref. 20) |
| PD ODE `d/dt(bcva) = kin·(1 + Emax_t·Cc/(EC50+Cc)) - kout·bcva` | n/a  | Eq. 1                                     |
| Time-dependent Emax `Emax_t = Emax_ss + dEmax_0·exp(-kEmax·t)` | n/a  | Eq. 2                                     |
| Covariate form `theta · (AGE/77)^beta_AGE · exp(eta)`         | n/a   | Eq. 3                                     |
| Natural-progression steady state `gss = kin/kout`            | n/a    | Methods — paragraph after Eq. 1           |
| `kel`        | 0.077/day (fixed; t½ = 9 d)          | Table 2 (borrowed from ref. 20)            |
| `V_vitreous` | 4 mL (fixed)                         | Methods — paragraph after Eq. 1           |
| `gss`        | 11 ETDRS letters                     | Table 2                                   |
| `kout`       | 0.19/year (t½ = 3.6 y)               | Table 2                                   |
| `Emax_ss`    | 6.1                                  | Table 2                                   |
| `dEmax_0`    | 41                                   | Table 2                                   |
| `kEmax`      | 0.046/day (fixed; t½ = 15 d)         | Table 2                                   |
| `EC50`       | 2.1 µg/mL                            | Table 2                                   |
| `β_Emax_ss,AGE` | −1.4                              | Table 2                                   |
| IIV `g0`     | SD 4.1 letters (additive)            | Table 2                                   |
| IIV `kout`   | CV 730 %                             | Table 2                                   |
| IIV `Emax_ss`| CV 110 %                             | Table 2                                   |
| IIV `dEmax_0`| CV 1100 %                            | Table 2                                   |
| σ treatment  | 5 letters (additive)                 | Table 2                                   |
| σ sham       | 7 letters (additive, not exposed here) | Table 2 (see Assumptions and deviations) |

Unit check at the reference dose: 0.5 mg delivered into 4 mL vitreous gives
peak concentration 0.5 / 0.004 = 125 mg/L ≡ 125 µg/mL. After 30 days,
$125 \cdot e^{-0.077 \cdot 30} = 12.4$ µg/mL (paper Methods text: "12.5 µg/mL
one month after an injection"); after 90 days, $125 \cdot e^{-0.077 \cdot 90}
= 0.12$ µg/mL (paper: "0.12 µg/mL three months after an injection"). The
implementation therefore matches the paper's published trough concentrations
to < 1 %.

# Virtual cohort

The original BCVA dataset is not publicly available. For validation we
construct a virtual cohort that matches the pooled baseline demographics
(mean age 77 years; mean baseline BCVA 54 letters) and also match the HARBOR
0.5 mg q4w cohort used as the paper's external validation (mean age 78.8
years; mean baseline BCVA 54.2 letters, Table 1). Figures 3 and 4 of the
paper depict mean predictions over simulated populations, so we run **both**
a typical-value (zero-IIV) trajectory and a stochastic cohort of 1,000
subjects and average.

```{r cohort}
set.seed(20260424)

n_cohort <- 1000L

# Pooled-demographics cohort (paper Results paragraph: mean age 77, mean baseline BCVA 54)
pooled_cohort <- tibble(
  id   = seq_len(n_cohort),
  AGE  = pmax(50, pmin(96, rnorm(n_cohort, mean = 77.0, sd = 7.5))),
  BCVA = pmax( 3, pmin(84, rnorm(n_cohort, mean = 54.0, sd = 13.0))),
  cohort = "Pooled (ANCHOR/MARINA/PIER/EXCITE)"
)

# HARBOR 0.5 mg q4w external-validation cohort (Table 1 column "HARBOR q4w 0.5 mg")
harbor_05_cohort <- tibble(
  id   = n_cohort + seq_len(n_cohort),
  AGE  = pmax(50, pmin(96, rnorm(n_cohort, mean = 78.8, sd = 8.4))),
  BCVA = pmax( 3, pmin(84, rnorm(n_cohort, mean = 54.2, sd = 13.3))),
  cohort = "HARBOR 0.5 mg q4w"
)

# HARBOR 2.0 mg q4w external-validation cohort (Table 1 column "HARBOR q4w 2.0 mg")
harbor_20_cohort <- tibble(
  id   = 2L * n_cohort + seq_len(n_cohort),
  AGE  = pmax(50, pmin(96, rnorm(n_cohort, mean = 79.3, sd = 8.3))),
  BCVA = pmax( 3, pmin(84, rnorm(n_cohort, mean = 53.5, sd = 13.1))),
  cohort = "HARBOR 2.0 mg q4w"
)

# Event-table helper: 12 monthly doses (day 0, 28, 56, ..., 308) + monthly BCVA obs.
make_events <- function(cov_df, dose_mg, obs_days = seq(0, 336, by = 7)) {
  ids  <- cov_df$id
  dose_times <- seq(0, by = 28, length.out = 12)
  dose_rows <- expand.grid(id = ids, time = dose_times) |>
    mutate(evid = 1L, amt = dose_mg, cmt = "central", dv = NA_real_) |>
    as_tibble()
  obs_rows <- expand.grid(id = ids, time = obs_days) |>
    mutate(evid = 0L, amt = 0, cmt = "bcva", dv = NA_real_) |>
    as_tibble()
  bind_rows(dose_rows, obs_rows) |>
    arrange(id, time, desc(evid)) |>
    left_join(cov_df, by = "id")
}

events_pooled    <- make_events(pooled_cohort,    dose_mg = 0.5)
events_harbor_05 <- make_events(harbor_05_cohort, dose_mg = 0.5)
events_harbor_20 <- make_events(harbor_20_cohort, dose_mg = 2.0)

stopifnot(!anyDuplicated(unique(
  bind_rows(events_pooled, events_harbor_05, events_harbor_20)[, c("id", "time", "evid")]
)))
```

# Simulation

```{r simulate}
mod         <- readModelDb("Mulyukov_2018_ranibizumab")
mod_typical <- rxode2::zeroRe(mod)

sim_typical_harbor_05 <- rxode2::rxSolve(
  mod_typical,
  events = events_harbor_05 |> filter(id == harbor_05_cohort$id[1]),
  returnType = "data.frame"
)

sim_harbor_05 <- rxode2::rxSolve(
  mod,
  events = events_harbor_05,
  keep   = c("cohort"),
  returnType = "data.frame"
)

sim_harbor_20 <- rxode2::rxSolve(
  mod,
  events = events_harbor_20,
  keep   = c("cohort"),
  returnType = "data.frame"
)

sim_pooled <- rxode2::rxSolve(
  mod,
  events = events_pooled,
  keep   = c("cohort"),
  returnType = "data.frame"
)
```

# Replicate published figures

## PK unit check — vitreous concentration after a single 0.5 mg injection

Mulyukov 2018 Methods cites two concrete PK landmarks from the borrowed
PK model: $C_{vitreous}(30\text{ d}) = 12.5$ µg/mL and
$C_{vitreous}(90\text{ d}) = 0.12$ µg/mL after a single 0.5 mg
intravitreal injection.

```{r pk-landmarks}
ev_single <- data.frame(
  id = 1L,
  time = c(0, 0.001, 30, 90),
  evid = c(1, 0, 0, 0),
  amt  = c(0.5, 0, 0, 0),
  cmt  = c("central", "bcva", "bcva", "bcva"),
  AGE = 77, BCVA = 55
)
sim_single <- rxode2::rxSolve(mod_typical, events = ev_single, returnType = "data.frame")
pk_check <- tibble(
  time           = c(0, 30, 90),
  Cc             = c(sim_single$Cc[which.min(abs(sim_single$time - 0.001))],
                     sim_single$Cc[which.min(abs(sim_single$time - 30))],
                     sim_single$Cc[which.min(abs(sim_single$time - 90))]),
  paper_Cc_ug_mL = c(125, 12.5, 0.12)
) |> mutate(pct_diff = 100 * (Cc - paper_Cc_ug_mL) / paper_Cc_ug_mL)
knitr::kable(pk_check,
  caption = "Vitreous concentration after a single 0.5 mg intravitreal injection, simulated vs. paper Methods.")
```

## Figure 2 analogue — typical-value BCVA trajectory under 0.5 mg q4w (HARBOR covariates)

```{r figure-2}
sim_pooled_summary <- sim_pooled |>
  group_by(time, cohort) |>
  summarise(
    bcva_mean = mean(bcva, na.rm = TRUE),
    bcva_Q05  = quantile(bcva, 0.05, na.rm = TRUE),
    bcva_Q50  = quantile(bcva, 0.50, na.rm = TRUE),
    bcva_Q95  = quantile(bcva, 0.95, na.rm = TRUE),
    .groups   = "drop"
  )

ggplot(sim_pooled_summary, aes(time, bcva_Q50)) +
  geom_ribbon(aes(ymin = bcva_Q05, ymax = bcva_Q95), alpha = 0.2) +
  geom_line(linewidth = 0.7) +
  geom_line(aes(y = bcva_mean), colour = "steelblue", linewidth = 0.7) +
  labs(
    x = "Time (day)", y = "BCVA (ETDRS letters)",
    title = "Simulated BCVA trajectory — ranibizumab 0.5 mg q4w, pooled-demographics cohort",
    caption = "Median (black) and 5th-95th percentile envelope from 1,000 simulated subjects; mean overlaid in blue. Replicates the treated-arm panels of Figure 2 of Mulyukov 2018."
  )
```

## Figure 3 — HARBOR external validation at 12 months

The paper reports (Results — Model evaluation) observed HARBOR 12-month mean
BCVA changes of approximately +10 and +9 ETDRS letters for the 0.5 mg and
2.0 mg q4w arms, with model-predicted +8.5 and +9.2 letters respectively.

```{r figure-3}
cfb_12mo <- bind_rows(
  sim_harbor_05 |>
    group_by(id, cohort) |>
    summarise(bcva_0 = first(bcva), bcva_12 = bcva[which.min(abs(time - 336))],
              .groups = "drop"),
  sim_harbor_20 |>
    group_by(id, cohort) |>
    summarise(bcva_0 = first(bcva), bcva_12 = bcva[which.min(abs(time - 336))],
              .groups = "drop")
) |>
  mutate(delta_bcva_12 = bcva_12 - bcva_0)

cfb_summary <- cfb_12mo |>
  group_by(cohort) |>
  summarise(
    mean_delta   = mean(delta_bcva_12),
    median_delta = median(delta_bcva_12),
    SD_delta     = sd(delta_bcva_12),
    n            = dplyr::n(),
    .groups      = "drop"
  ) |>
  mutate(
    paper_observed_letters  = c(10.1, 9.2),
    paper_predicted_letters = c( 8.5, 9.2)
  )

knitr::kable(cfb_summary, digits = 2,
  caption = "Simulated vs. published BCVA change from baseline at 12 months (ranibizumab HARBOR arms). Paper values from Results — Model evaluation and simulations and Table 1.")
```

## Figure 4a analogue — dependence of 12-month BCVA change on baseline BCVA

Replicates Figure 4a of Mulyukov 2018: mean 12-month BCVA change decreases
linearly with baseline BCVA ("for every 10 letters of lower baseline BCVA
there are 3 letter gains in BCVA improvement").

```{r figure-4a}
cfb_baseline <- sim_pooled |>
  group_by(id) |>
  summarise(bcva_0 = first(bcva), bcva_12 = bcva[which.min(abs(time - 336))],
            BCVA   = first(BCVA), AGE = first(AGE), .groups = "drop") |>
  mutate(delta_bcva_12 = bcva_12 - bcva_0)

bcva_bins <- cfb_baseline |>
  mutate(bcva_bin = cut(BCVA, breaks = seq(0, 90, by = 10), include.lowest = TRUE)) |>
  group_by(bcva_bin) |>
  summarise(mean_BCVA = mean(BCVA), mean_delta = mean(delta_bcva_12),
            n = dplyr::n(), .groups = "drop") |>
  filter(n >= 5)

ggplot(cfb_baseline, aes(BCVA, delta_bcva_12)) +
  geom_point(alpha = 0.12, size = 0.6) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, colour = "steelblue") +
  geom_point(data = bcva_bins, aes(mean_BCVA, mean_delta),
             colour = "firebrick", size = 2.5) +
  labs(
    x = "Baseline BCVA (ETDRS letters)",
    y = "BCVA change at month 12 (ETDRS letters)",
    title = "Figure 4a analogue — BCVA change vs. baseline BCVA (pooled cohort)",
    caption = "Blue line = linear fit. Red points = per-10-letter-bin means. Replicates Figure 4a of Mulyukov 2018."
  )
```

## Figure 4b analogue — dependence of 12-month BCVA change on age

Replicates Figure 4b of Mulyukov 2018: older patients show smaller BCVA gains
(≈ 4-letter reduction from age 65 to 85 at the reference baseline BCVA).

```{r figure-4b}
age_bins <- cfb_baseline |>
  mutate(age_bin = cut(AGE, breaks = seq(50, 100, by = 5), include.lowest = TRUE)) |>
  group_by(age_bin) |>
  summarise(mean_AGE = mean(AGE), mean_delta = mean(delta_bcva_12),
            n = dplyr::n(), .groups = "drop") |>
  filter(n >= 5)

ggplot(cfb_baseline, aes(AGE, delta_bcva_12)) +
  geom_point(alpha = 0.12, size = 0.6) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, colour = "steelblue") +
  geom_point(data = age_bins, aes(mean_AGE, mean_delta),
             colour = "firebrick", size = 2.5) +
  labs(
    x = "Age (years)",
    y = "BCVA change at month 12 (ETDRS letters)",
    title = "Figure 4b analogue — BCVA change vs. age (pooled cohort)",
    caption = "Blue line = linear fit. Red points = per-5-year-bin means. Replicates Figure 4b of Mulyukov 2018."
  )
```

## Natural-progression check — untreated decay toward `gss`

No-dose simulation confirms the Methods-stated behaviour: BCVA decays from
the observed baseline toward the steady-state value $g_{ss} \approx 11$
ETDRS letters at rate $k_{out} = 0.19$/year ($t_{1/2}$ ≈ 3.6 y). With no
treatment, g(1 year) should lose ~20 % of (baseline − gss).

```{r natural-progression}
ev_natural <- data.frame(
  id = 1L, time = seq(0, 365*4, length.out = 100),
  evid = 0L, amt = 0, cmt = "bcva",
  AGE = 77, BCVA = 55
)
sim_natural <- rxode2::rxSolve(mod_typical, events = ev_natural, returnType = "data.frame")
sim_natural |>
  filter(time %in% c(0, 365, 365*2, 365*3, 365*4)) |>
  transmute(year = round(time / 365, 1), bcva_predicted = round(bcva, 2),
            expected_decay_fraction = round(1 - exp(-0.19 * (time/365)), 3)) |>
  knitr::kable(caption = "Untreated BCVA trajectory from baseline 55 toward gss = 11 letters.")
```

# PKNCA validation

The borrowed PK model is a single-compartment first-order decay with a
reference t½ of 9 days; PKNCA confirms the implementation produces the
expected half-life and AUC for a single 0.5 mg intravitreal injection. Per
the skill's guidance, the formula includes a treatment grouping variable
(`dose_group`) so the result rolls up one row per regimen.

```{r pknca}
ev_nca <- bind_rows(
  data.frame(
    id = 1L, time = c(0, seq(0.01, 120, length.out = 200)),
    evid = c(1, rep(0, 200)),
    amt = c(0.5, rep(0, 200)),
    cmt = c("central", rep("bcva", 200)),
    AGE = 77, BCVA = 55,
    dose_group = "0.5 mg IVT single"
  ),
  data.frame(
    id = 2L, time = c(0, seq(0.01, 120, length.out = 200)),
    evid = c(1, rep(0, 200)),
    amt = c(2.0, rep(0, 200)),
    cmt = c("central", rep("bcva", 200)),
    AGE = 77, BCVA = 55,
    dose_group = "2.0 mg IVT single"
  )
)

sim_nca <- rxode2::rxSolve(mod_typical, events = ev_nca,
                            keep = c("dose_group"),
                            returnType = "data.frame") |>
  filter(time > 0)

conc_obj <- PKNCA::PKNCAconc(sim_nca,
  Cc ~ time | dose_group + id)
dose_df <- ev_nca |> filter(evid == 1) |>
  transmute(id, time, amt, dose_group)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | dose_group + id)

intervals <- data.frame(
  start         = 0,
  end           = 120,
  cmax          = TRUE,
  tmax          = TRUE,
  auclast       = TRUE,
  aucinf.obs    = TRUE,
  half.life     = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- PKNCA::pk.nca(nca_data)
nca_summary <- summary(nca_res, drop.group = "id")
nca_summary
```

## Comparison against published PK landmarks

The paper does not publish a full NCA table. Two PK landmarks are
verifiable against the simulation:

- $C_{vitreous}(1\text{ month})$: paper 12.5 µg/mL vs. simulated
  `r sprintf("%.2f", sim_nca$Cc[which.min(abs(sim_nca$time - 30) + 1e4*(sim_nca$dose_group != "0.5 mg IVT single"))])` µg/mL.
- $C_{vitreous}(3\text{ months})$: paper 0.12 µg/mL vs. simulated
  `r sprintf("%.3f", sim_nca$Cc[which.min(abs(sim_nca$time - 90) + 1e4*(sim_nca$dose_group != "0.5 mg IVT single"))])` µg/mL.
- Half-life: paper 9 days; PKNCA returns the value shown in the summary
  above.

All three match the paper to < 1 %.

# Assumptions and deviations

- **Single residual-error SD is exposed.** The paper reports two additive
  residual-error SDs on BCVA in Table 2 — $\sigma_{treatment} = 5$ letters for
  treated arms and $\sigma_{sham} = 7$ letters for untreated arms. This
  library model exposes only $\sigma_{treatment} = 5$ letters (the primary
  use case for ranibizumab simulation is the treated-arm trajectory). If a
  user needs to simulate a sham/untreated arm with the paper's residual
  error, override `bcvaaddSd` to `7` after loading the model. The typical-
  value and individual structural trajectories are unchanged.
- **Baseline BCVA treated as a covariate.** The paper writes
  $g_{i,0} = BVA_i + \eta_{1,i}$ where $BVA_i$ is the observed baseline VA
  for subject $i$. We expose $BVA_i$ as the canonical covariate
  column `BCVA` and pair it with an additive typical-value theta
  `g0res = fixed(0)` so the eta pairing satisfies the `eta<x>`-pairs-with-
  `<x>` library convention. Users must supply per-subject `BCVA` at the
  time of the first dose (time 0).
- **`kEmax` taken from Table 2 (0.046/day; t½ = 15 days), not from the
  Methods text (log(2)/14 = 0.0495/day; t½ = 14 days).** Table 2 presents
  the final-model estimates, so it supersedes the text wording. The paper
  states the model is insensitive to any $k_{Emax}$ corresponding to a
  1-to-3-week Emax half-life, so the discrepancy between the two published
  values is not operationally meaningful.
- **Log-normal mean ≠ typical value.** The paper reports extremely large
  CVs on the PD parameters (CV% 730 for $k_{out}$, 1100 for $\Delta E_{\max}^0$).
  Because $E[\exp(\eta)] = \exp(\omega^2/2)$, the population mean of a
  simulation with full IIV is materially larger than the typical-value
  trajectory. The mean BCVA-change curve in a 1,000-subject simulation of
  the HARBOR 0.5 mg arm is what the paper's Figure 3 depicts, not the
  typical-value trajectory. The `cohort` summary table above includes the
  mean delta alongside the paper's predicted and observed values.
- **Two `checkModelConventions()` warnings accepted.** (1) The compartment
  is named `bcva` rather than `effect`, matching the precedent of
  `Ma_2020_sarilumab_das28crp` (`das28`) and `Valenzuela_2025_nipocalimab`
  (named PD states) for clarity. (2) The single observation variable is
  `bcva` rather than `Cc`, because the observation is a visual-acuity score
  in ETDRS letters, not a concentration — the `Cc` convention (concentration
  in the central compartment) does not fit.
- **Race-ethnicity and body-weight distributions not simulated.** Table 1
  of the paper does not report race or body-weight breakdowns, so the
  virtual cohorts use only age and baseline BCVA. This omission does not
  affect model structure (no race / weight covariates were retained).
- **No errata found.** A search of the journal landing page and a
  PubMed/Google Scholar query ("Mulyukov 2018 ranibizumab erratum")
  returned no corrections. Paper values are the authoritative source.
