---
title: "Tovecimig / ABL001 / CTX-009 (Na 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tovecimig / ABL001 / CTX-009 (Na 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Na_2024_tovecimig"))$reference`
* Description: `r rxode2::rxode(readModelDb("Na_2024_tovecimig"))$description`
* Article: <https://doi.org/10.1111/cas.16363>
* Supplement (Figures S1-S5, Table S1): available from the article's Supporting
  Information section.

Tovecimig (development codes ABL001 and CTX-009) is a bispecific antibody that
simultaneously blocks delta-like ligand-4 (DLL4) and vascular endothelial growth
factor A (VEGF-A). Na 2024 developed the population PK model from the
dose-escalation portion of the first-in-human study NCT03292783.

# Population

The model was built on 712 plasma concentrations from 31 adults with progressive
locally advanced or metastatic solid tumors that had relapsed on, or were
refractory to, standard therapy (Na 2024 Table 1, dose-escalation cohort). The
14 patients in the dose-expansion cohort were *not* used for model development.
Patients were 25-81 years old (mean 53.5, SD 12.4), weighed 35.6-110.2 kg (mean
64.2, SD 16.4), and were 38.7% female. Nearly all (87.1%) had an ECOG
performance status of 1. The most common primary tumor sites were stomach
(38.7%) and colon/rectum (35.5%), and the median number of prior lines of
treatment was 5. All three study sites were in the Republic of Korea; race and
ethnicity were not tabulated.

Doses were 1-hour intravenous infusions of 0.3, 1, 2.5, 5, 7.5, 10, 12.5, 15, or
17.5 mg/kg given in 2-week cycles, except that the second dose followed the
first by 3 weeks. Only 6 of the 712 samples (< 1%) fell below the 250 ng/mL
lower limit of quantification and were treated as missing.

The same information is available programmatically via the model's `population`
metadata:

```{r population}
str(rxode2::rxode(readModelDb("Na_2024_tovecimig"))$population)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Na_2024_tovecimig.R`. The table below
collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL) | 0.0184 L/h | Na 2024 Table 2 (RSE 18%) |
| `lvc` (V1 at 70 kg) | 3.87 L | Na 2024 Table 2 (RSE 5%) |
| `lq` (Q) | 0.0207 L/h | Na 2024 Table 2 (RSE 16%) |
| `lvp` (V2) | 1.31 L | Na 2024 Table 2 (RSE 28%) |
| `lvmax` (Vmax) | 0.0968 mg/h | Na 2024 Table 2 (RSE 26%); printed unit "ug/h" is a mislabel, see Errata |
| `lkm` (Km, fixed) | 4.15 mg/L | Na 2024 Table 2 ("fixed"); value estimated in exploration model 4, Table S1; printed unit "ug/L" is a mislabel, see Errata |
| `e_wt_vc` (WT exponent on V1) | 0.598 | Na 2024 Table 2 (`theta_weight,V1`, RSE 32%) and Table 2 footnote b |
| `etalcl` | 0.258394 | Na 2024 Table 2, IIV CL 54.3 %CV; `log(0.543^2 + 1)` |
| `etalvc` | 0.024967 | Na 2024 Table 2, IIV V1 15.9 %CV; `log(0.159^2 + 1)` |
| `etalvp` | 0.699147 | Na 2024 Table 2, IIV V2 100.6 %CV; `log(1.006^2 + 1)` |
| `propSd` | 0.245 | Na 2024 Table 2 (RSE 1%), read as an SD; see Errata |
| Two-compartment structure with parallel linear and Michaelis-Menten elimination from the central compartment | n/a | Na 2024 Figure 1 and Section 3.2 |
| `V1 = theta_V1 * (WT / 70)^theta_weight,V1` | n/a | Na 2024 Section 2.4 Equation 1 and Table 2 footnote b |
| No IIV on Q or Vmax | n/a | Na 2024 Section 3.2 ("IIV for Vmax did not significantly improve the model and was not estimated") |
| Proportional-only residual error | n/a | Na 2024 Section 3.2 |
| Reference weight 70 kg | n/a | Na 2024 Section 2.4 ("a typically accepted value of 70 kg") |

# Typical-value profiles across the dose-escalation range

Because a saturable elimination pathway runs in parallel with linear clearance,
the model is dose-nonlinear: the Michaelis-Menten arm contributes a clearance of
`Vmax / (Km + Cc)`, which is largest at low concentrations. The panel below
reproduces the shape of the individual concentration-time profiles in Na 2024
Figure S2 for a typical 70 kg patient at each dose-escalation level.

The random effects are suppressed by supplying zero `eta` columns and
`omega = NA`, which gives a deterministic typical-value prediction without
mutating the shared model object.

```{r typical-profiles}
dose_levels <- c(0.3, 1, 2.5, 5, 7.5, 10, 12.5, 15, 17.5)
ref_wt <- 70

obs_grid <- c(seq(0, 24, by = 0.5), seq(25, 168, by = 1), seq(172, 1344, by = 4))

make_single_dose <- function(dose_mg_kg, wt, id) {
  amt <- dose_mg_kg * wt
  dplyr::bind_rows(
    tibble::tibble(id = id, time = 0, amt = amt, rate = amt, evid = 1L,
                   cmt = "central"),
    tibble::tibble(id = id, time = obs_grid, amt = NA_real_, rate = NA_real_,
                   evid = 0L, cmt = "central")
  ) |>
    dplyr::mutate(
      WT = wt,
      dose_label = sprintf("%s mg/kg", dose_mg_kg),
      etalcl = 0, etalvc = 0, etalvp = 0
    )
}

ev_typical <- dplyr::bind_rows(
  lapply(seq_along(dose_levels), function(i) {
    make_single_dose(dose_levels[i], ref_wt, id = i)
  })
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(!anyDuplicated(unique(ev_typical[, c("id", "time", "evid")])))

mod <- readModelDb("Na_2024_tovecimig")
sim_typical <- rxode2::rxSolve(
  mod, events = ev_typical, omega = NA,
  keep = c("WT", "dose_label"), addDosing = FALSE
) |>
  as.data.frame()

stopifnot(all(sim_typical$Cc >= 0), !anyNA(sim_typical$Cc))
```

```{r figure-s2, fig.cap = "Typical-value single-dose plasma profiles. Reproduces the shape of Figure S2 of Na 2024."}
lloq <- 0.25 # mg/L, Na 2024 Section 2.3 (250 ng/mL)

sim_typical |>
  dplyr::mutate(
    dose_label = factor(dose_label, levels = sprintf("%s mg/kg", dose_levels))
  ) |>
  dplyr::filter(Cc >= lloq) |>
  ggplot(aes(time / 24, Cc, colour = dose_label)) +
  geom_line() +
  geom_hline(yintercept = lloq, linetype = "dashed") +
  scale_y_log10() +
  labs(
    x = "Time since dose (days)", y = "Tovecimig plasma concentration (mg/L)",
    colour = "Dose",
    title = "Typical-value single-dose profiles (70 kg)",
    caption = "Dashed line: 250 ng/mL LLOQ. Replicates the shape of Figure S2 of Na 2024."
  )
```

The curvature on the semi-log scale is the signature of the saturable arm: the
terminal slope steepens as concentrations fall toward `Km`.

```{r halflife-check}
# Model-derived elimination half-life. Na 2024 (Discussion) reports
# "approximately 5.7 days". Because a Michaelis-Menten arm runs in parallel with
# linear CL, the effective half-life depends on concentration; evaluate it at the
# published steady-state average concentration for the 10 mg/kg Q2W regimen.
theta <- rxode2::rxode(readModelDb("Na_2024_tovecimig"))$theta
cl_lin <- exp(theta[["lcl"]])
v1     <- exp(theta[["lvc"]])
vmax   <- exp(theta[["lvmax"]])
km     <- exp(theta[["lkm"]])

cav_pub <- 111.3 # mg/L, Na 2024 Table 3 (10 mg/kg Q2W)
cl_eff  <- cl_lin + vmax / (km + cav_pub)
t_half_h <- log(2) * v1 / cl_eff

c(
  `CL linear (L/h)`          = cl_lin,
  `CL Michaelis-Menten (L/h)` = vmax / (km + cav_pub),
  `CL effective (L/h)`       = cl_eff,
  `t1/2 (days)`              = t_half_h / 24,
  `published t1/2 (days)`    = 5.7
) |>
  round(4)
```

The effective half-life computed from the packaged parameters is within 2% of
the 5.7 days the paper reports. Note that with the printed `ug/h` reading of
`Vmax` the Michaelis-Menten arm would be roughly 1000-fold smaller, the
effective half-life would be 6.1 days, and the steady-state exposures below
would be ~6% high; see Errata.

# Steady-state regimen comparison (Table 3, Figures 3 and 4)

Na 2024 simulated 1000 virtual patients receiving either a 700 mg fixed dose or
a 10 mg/kg weight-based dose, and separately compared 10 mg/kg every 2 weeks
against 15 mg/kg every 3 weeks. Body weights were drawn from the observed
patient distribution: mean 65.5 kg, SD 18.1 kg, spanning roughly 35-120 kg
(Na 2024 Section 2.6 and Figure S1).

The cohort here uses 200 subjects per arm (the nlmixr2lib cap) rather than the
paper's 1000. Two variance-reduction choices make that sufficient and make the
between-regimen comparisons sharp:

* **One shared cohort across all three arms.** The paper's own design gives the
  *same* virtual patients both the fixed and the weight-based regimen, so body
  weights and the three `eta` draws are generated once and reused for every arm
  (with disjoint `id` offsets). Between-arm differences are then entirely due to
  the regimen, not to resampling.
* **Stratified (quantile) sampling of the body weights.** The weight
  distribution is a fact about the trial population, so placing the `i`-th
  subject at the `(i - 0.5)/n` quantile reproduces it almost exactly at n = 200.
  The `eta` values are *not* treated this way: they are the model's own
  stochastic element, and pinning each one to its marginal quantiles would
  understate the joint tail of any quantity that depends on more than one of
  them. They are drawn randomly under a fixed seed and supplied as data columns
  with `omega = NA` so that all three arms share them.

Body weights follow the observed patient distribution: mean 65.5 kg, SD 18.1 kg,
spanning roughly 35-120 kg (Na 2024 Section 2.6 and Figure S1). Because the
reported mean describes patients who all lie inside the 35-120 kg range, the
underlying normal is centered so that the *truncated* distribution has mean
65.5 kg.

```{r cohort}
set.seed(20241203)

n_per_arm <- 200L
tau_q2w <- 336 # h
tau_q3w <- 504 # h
wt_sd <- 18.1
wt_lo <- 35
wt_hi <- 120

# Center the underlying normal so the truncated [35, 120] distribution has the
# reported mean of 65.5 kg.
truncated_mean <- function(mu) {
  a <- (wt_lo - mu) / wt_sd
  b <- (wt_hi - mu) / wt_sd
  mu + wt_sd * (stats::dnorm(a) - stats::dnorm(b)) /
    (stats::pnorm(b) - stats::pnorm(a))
}
mu_wt <- stats::uniroot(function(m) truncated_mean(m) - 65.5, c(40, 95))$root

# Stratified draw from the truncated normal.
probs <- (seq_len(n_per_arm) - 0.5) / n_per_arm
p_lo <- stats::pnorm(wt_lo, mu_wt, wt_sd)
p_hi <- stats::pnorm(wt_hi, mu_wt, wt_sd)
wt <- sample(stats::qnorm(p_lo + probs * (p_hi - p_lo), mu_wt, wt_sd))

# Random eta draws (NOT quantile-pinned -- see the note above), using the omega
# variances carried by the model itself.
ui <- rxode2::rxode(readModelDb("Na_2024_tovecimig"))
omega_var <- diag(ui$omega)
draw_eta <- function(nm) stats::rnorm(n_per_arm, 0, sqrt(omega_var[[nm]]))

subj_base <- tibble::tibble(
  subj   = seq_len(n_per_arm),
  WT     = wt,
  etalcl = draw_eta("etalcl"),
  etalvc = draw_eta("etalvc"),
  etalvp = draw_eta("etalvp")
)

c(`cohort mean WT (kg)` = mean(wt), `cohort SD WT (kg)` = stats::sd(wt),
  `min WT` = min(wt), `max WT` = max(wt)) |>
  round(1)

# Both q2w and q3w arms give their final dose at the same absolute time
# (10080 h, 60 weeks), which is many effective half-lives and reaches steady
# state even for the subjects with the largest peripheral volume.
t_last_dose <- 10080

make_arm <- function(regimen, dose_fun, tau, id_offset) {
  n_doses <- t_last_dose / tau + 1
  subj <- subj_base |>
    dplyr::mutate(
      id = id_offset + subj,
      amt_mg = dose_fun(WT),
      regimen = regimen
    )
  dosing <- subj |>
    tidyr::crossing(time = seq(0, by = tau, length.out = n_doses)) |>
    dplyr::mutate(amt = amt_mg, rate = amt_mg, evid = 1L, cmt = "central")
  # Dense sampling over the final (steady-state) dosing interval only.
  interval_grid <- c(seq(0, 24, by = 1), seq(26, tau, by = 6))
  if (!tau %in% interval_grid) interval_grid <- c(interval_grid, tau)
  obs <- subj |>
    tidyr::crossing(time = t_last_dose + sort(unique(interval_grid))) |>
    dplyr::mutate(amt = NA_real_, rate = NA_real_, evid = 0L, cmt = "central")
  dplyr::bind_rows(dosing, obs) |>
    dplyr::select(id, time, amt, rate, evid, cmt, WT, regimen,
                  etalcl, etalvc, etalvp) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  make_arm("700 mg fixed Q2W", function(w) rep(700, length(w)), tau_q2w,   0L),
  make_arm("10 mg/kg Q2W",     function(w) 10 * w,              tau_q2w, 200L),
  make_arm("15 mg/kg Q3W",     function(w) 15 * w,              tau_q3w, 400L)
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

```{r simulate}
mod_ss <- readModelDb("Na_2024_tovecimig")
sim <- rxode2::rxSolve(
  mod_ss, events = events, omega = NA,
  keep = c("WT", "regimen"), addDosing = FALSE
) |>
  as.data.frame()

stopifnot(all(sim$Cc >= 0), !anyNA(sim$Cc))
```

`Cc` is the individual predicted concentration; the published simulations do not
appear to include residual error (see Errata), so `Cc` rather than the
residual-error-perturbed `sim` column is the like-for-like quantity.

```{r figure-3, fig.cap = "Steady-state concentration-time profiles over the final dosing interval. Replicates Figure 3 of Na 2024."}
sim |>
  dplyr::mutate(tad = time - t_last_dose) |>
  dplyr::group_by(regimen, tad) |>
  dplyr::summarise(
    Q05 = stats::quantile(Cc, 0.05),
    Q50 = stats::median(Cc),
    Q95 = stats::quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad / 24, Q50, colour = regimen, fill = regimen)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.15, colour = NA) +
  geom_line() +
  labs(
    x = "Time after the final dose (days)",
    y = "Tovecimig plasma concentration (mg/L)",
    colour = "Regimen", fill = "Regimen",
    title = "Steady-state profiles: median with 5th-95th percentiles",
    caption = "Replicates Figure 3 of Na 2024."
  )
```

```{r figure-4, fig.cap = "Distribution of the steady-state average concentration by regimen. Replicates Figure 4 of Na 2024."}
cav_by_id <- sim |>
  dplyr::group_by(id, regimen) |>
  dplyr::arrange(time, .by_group = TRUE) |>
  dplyr::summarise(
    tau_h = max(time) - min(time),
    auc = sum(diff(time) * (utils::head(Cc, -1) + utils::tail(Cc, -1)) / 2),
    .groups = "drop"
  ) |>
  dplyr::mutate(cav = auc / tau_h)

ggplot(cav_by_id, aes(regimen, cav, fill = regimen)) +
  geom_boxplot(alpha = 0.5, outlier.alpha = 0.3) +
  labs(
    x = NULL, y = "Steady-state average concentration (mg/L)",
    title = "C average at steady state by regimen",
    caption = "Replicates Figure 4 of Na 2024."
  ) +
  theme(legend.position = "none")
```

# PKNCA validation

Na 2024 Table 3 reports a predose trough concentration. This is an intravenous
model with no absorption phase, so within a steady-state dosing interval the
concentration falls monotonically after the end of infusion and the interval
minimum *is* the predose trough. PKNCA's `cmin` is therefore the like-for-like
statistic; `ctrough` cannot be used here because PKNCA compares dose-relative
times against an absolute interval end. The check below confirms the
identification rather than assuming it.

```{r trough-identity}
trough_check <- sim |>
  dplyr::group_by(id, regimen) |>
  dplyr::arrange(time, .by_group = TRUE) |>
  dplyr::summarise(
    c_start = dplyr::first(Cc),
    c_end   = dplyr::last(Cc),
    c_min   = min(Cc),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    ss_rel_diff   = abs(c_end - c_start) / c_start,
    min_vs_trough = abs(c_min - c_end) / c_end
  )

# The interval opens on the pre-dose trough and closes on the next pre-dose
# trough, so at steady state those two are the same number and the interval
# minimum equals both.
stopifnot(
  nrow(trough_check) == 3L * n_per_arm,
  max(trough_check$ss_rel_diff)   < 1e-3, # steady state reached
  max(trough_check$min_vs_trough) < 1e-3  # interval minimum IS the trough
)

c(`max |C(end) - C(start)| / C(start)` = max(trough_check$ss_rel_diff),
  `max |Cmin - C(end)| / C(end)`       = max(trough_check$min_vs_trough)) |>
  signif(3)
```

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, regimen)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, regimen)

conc_obj <- PKNCA::PKNCAconc(
  sim_nca, Cc ~ time | regimen + id,
  concu = "mg/L", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  dose_df, amt ~ time | regimen + id,
  doseu = "mg"
)

# Steady-state interval = the final dosing interval of each regimen.
intervals <- data.frame(
  regimen = c("700 mg fixed Q2W", "10 mg/kg Q2W", "15 mg/kg Q3W"),
  start   = t_last_dose,
  end     = t_last_dose + c(tau_q2w, tau_q2w, tau_q3w),
  cmax    = TRUE,
  tmax    = TRUE,
  auclast = TRUE,
  cav     = TRUE,
  cmin    = TRUE,
  stringsAsFactors = FALSE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)
```

```{r pknca-summary}
nca_wide <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "auclast", "cav", "cmin")) |>
  dplyr::group_by(regimen, PPTESTCD) |>
  dplyr::summarise(
    mean = mean(PPORRES), sd = stats::sd(PPORRES), .groups = "drop"
  )

nca_wide |>
  dplyr::mutate(
    value = sprintf("%.1f +/- %.1f (SEM %.1f%%)", mean, sd,
                    100 * sd / sqrt(n_per_arm) / mean),
    PPTESTCD = nlmixr2lib::ncaParamLabel(PPTESTCD)
  ) |>
  dplyr::select(regimen, PPTESTCD, value) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = value) |>
  dplyr::rename("Regimen" = regimen) |>
  knitr::kable(
    caption = paste(
      "Simulated steady-state exposure, mean +/- SD across 200 virtual",
      "patients per arm, with the Monte-Carlo standard error of the mean.",
      "Compare with Table 3 of Na 2024."
    )
  )
```

## Comparison against published NCA

Na 2024 Table 3 reports arithmetic means, so the simulated side is summarised as
an arithmetic mean per arm (rather than the median `ncaComparisonTable()` would
take from a raw `PKNCAresults` object) to keep the comparison like-for-like.

```{r nca_compare}
simulated_means <- nca_wide |>
  dplyr::select(regimen, PPTESTCD, mean) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = mean)

published <- tibble::tribble(
  ~regimen,            ~cmax,  ~auclast, ~cav,   ~cmin,
  "700 mg fixed Q2W",  263.1,  41638.5,  123.9,  66.5,
  "10 mg/kg Q2W",      232.3,  37407.2,  111.3,  60.4,
  "15 mg/kg Q3W",      296.1,  54255.6,  107.6,  45.4
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = simulated_means,
  reference = published,
  by        = "regimen",
  units     = c(cmax = "mg/L", auclast = "h*mg/L",
                cav = "mg/L", cmin = "mg/L"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated vs. published steady-state NCA (Na 2024 Table 3).",
    "* differs from the reference by more than 20%."
  )
)
```

The 700 mg fixed-dose arm reproduces all four published quantities to within
2.1%, and its AUC and Cavg to within 0.3%. That is a strong joint check: Cmax
pins `V1` and its weight exponent, AUC over the dosing interval pins total
clearance (hence both the linear `CL` and the Michaelis-Menten arm, and with it
the `mg` reading of `Vmax` and `Km`), and the standard deviations reported
alongside them pin the three `omega` terms -- the simulated SDs sit within a few
percent of the published ones for every quantity.

The weight-based arms sit 0.8-8.3% high. Most of that is inside the Monte-Carlo
noise of a 200-subject cohort: the standard error of the mean printed in the
table above is 3.6-3.9% for AUC and Cavg and about 6% for Cmin, so only the
15 mg/kg Q3W AUC sits as much as two standard errors away. Part of the rest is
arithmetic in the
paper's own table rather than in this implementation. At steady state
AUC over a dosing interval is dose divided by clearance, so the 15 mg/kg Q3W and
10 mg/kg Q2W rows should differ by the dose ratio 1.500 in a linear model, and
by slightly less once saturable elimination is allowed. Na 2024 reports
54255.6 / 37407.2 = 1.450, a 3.3% shortfall. For this model the largest possible
shortfall is bounded by `(Km + 111.3) / (Km + 107.6) = 1.033`, and that bound is
only attained if the linear clearance were zero; with the published
`CL = 0.0184 L/h` the achievable sublinearity is under 0.1%. So the two
weight-based rows of Table 3 cannot both be reproduced by any parameterization
of the paper's own model, and the residual is most plausibly Monte-Carlo noise
in the published simulation (with a 59% coefficient of variation on AUC and
1000 virtual patients, the standard error on each published mean is about 1.9%).

Every row is well inside the 20% tolerance and none is flagged. No parameter was
adjusted.

# Assumptions and deviations

## Errata and source conflicts

* **`Vmax` and `Km` units.** Na 2024 Table 2 prints `Vmax (ug/h) = 0.0968` and
  `Km (ug/L) = 4.15`. Taken literally, the saturable arm would contribute a
  clearance of about `1e-6` L/h, i.e. nothing at all, which contradicts the
  paper elsewhere. Three arguments select the `mg` reading, the first of which
  is decisive on its own.

  (a) *Dimensional analysis.* The concentration data are in mg/L throughout
  (Table 3 reports Cmax and Ctrough in mg/L) and doses are in mg. A NONMEM
  `Vmax` is an amount per time in the dataset's own units, so it is mg/h, and
  `Km` is a concentration in the dataset's own units, so it is mg/L.

  (b) *The reported half-life.* "Approximately 5.7 days" requires the extra
  clearance; the `mg` reading gives 5.78 days and the `ug` reading 6.1 days, as
  the closed-form check earlier in this vignette shows.

  (c) *Consistency with Table 3.* The `mg` reading reproduces the 700 mg
  fixed-dose AUC over the dosing interval to +0.3%. Because that arm's AUC is
  dose divided by clearance, and is therefore insensitive both to `V1` and to
  the cohort's weights, it is a clean check on total clearance. The `ug` reading
  removes the Michaelis-Menten contribution entirely and would raise it by
  roughly 6%.

  The model file therefore uses `Vmax = 0.0968 mg/h` and `Km = 4.15 mg/L`, and
  the `ug` labels are treated as printing errors. No value was changed, only its
  unit interpretation.
* **`theta_V1`: 3.87 L (Table 2) vs 3.64 L (Results equation).** The Results
  text prints `V1 (L) = 3.64 (L) x (Weight (kg) / 70 (kg))^0.598`, but Table 2
  reports the final estimate as 3.87 L (RSE 5%) with a bootstrap median of
  3.63 L. The model file uses the Table 2 value, 3.87 L, on the strength of the
  table itself: Table 2 is the paper's authoritative "Parameter estimates of the
  final model" and carries an RSE, whereas the 3.64 printed in the Results
  sentence coincides with the bootstrap median 3.63 and reads as a transcription
  slip from the adjacent column.

  Simulation does **not** settle this one, and it is worth showing why rather
  than implying otherwise. In this model `V1` enters only as
  `exp(lvc + etalvc) * (WT/70)^0.598`, so substituting 3.64 L for 3.87 L is
  exactly a constant shift of `log(3.64/3.87)` on `etalvc`; the check below
  re-solves the identical cohort with that shift, so the two readings differ in
  nothing else.

```{r v1-sensitivity}
sim_v1_alt <- rxode2::rxSolve(
  mod_ss,
  events = events |> dplyr::mutate(etalvc = etalvc + log(3.64 / 3.87)),
  omega = NA, keep = c("WT", "regimen"), addDosing = FALSE
) |>
  as.data.frame()

fixed_arm_cmax <- function(x) {
  x |>
    dplyr::filter(regimen == "700 mg fixed Q2W") |>
    dplyr::group_by(id) |>
    dplyr::summarise(cmax = max(Cc), .groups = "drop") |>
    dplyr::pull(cmax) |>
    mean()
}

c(`published Cmax (mg/L)`      = 263.1,
  `theta_V1 = 3.87 (Table 2)`  = fixed_arm_cmax(sim),
  `theta_V1 = 3.64 (Results)`  = fixed_arm_cmax(sim_v1_alt)) |>
  round(1)
```

  The two readings straddle the published mean steady-state Cmax by a similar
  margin (-2.1% and +1.7%), and both sit inside roughly one Monte-Carlo standard
  error of this cohort's mean Cmax. Cmax is also sensitive to the cohort's
  weight distribution, since `V1` scales as `WT^0.598`. Table 3 therefore cannot
  discriminate the two values at the available precision; the choice rests on
  Table 2's authority, and the check above serves to bound the consequence of
  the alternative reading at about 4% in simulated Cmax rather than to justify
  the choice.
* **Residual error, SD vs variance.** Table 2 reports "Proportional residual
  error 0.245" with no unit or scale annotation. It is encoded as an SD
  (`propSd = 0.245`, i.e. 24.5% CV). Read as a NONMEM variance it would imply a
  49.5% CV proportional error, which is inconsistent with the assay precision
  reported in Section 2.3 (accuracy within -1.60% to 3.76%, precision below
  12.76%) and with the tightness of the pcVPC in Figure 2. Na 2024's Table 3
  simulations do not include residual error at all (see the next bullet), so
  this choice does not affect any comparison in this vignette.
* **Table 3 was simulated without residual error.** Cmax is a maximum over a
  dense sampling grid, so it is strongly inflated by any residual error, while
  AUC over the interval is barely affected. Applying the model's proportional
  error to the same simulation shows the published Cmax is only compatible with
  a residual-error-free simulation:

```{r ruv-check}
set.seed(20241205)
prop_sd <- rxode2::rxode(readModelDb("Na_2024_tovecimig"))$theta[["propSd"]]

sim |>
  dplyr::filter(regimen == "700 mg fixed Q2W") |>
  dplyr::mutate(dv = Cc * (1 + stats::rnorm(dplyr::n(), 0, prop_sd))) |>
  dplyr::group_by(id) |>
  dplyr::summarise(cmax_ipred = max(Cc), cmax_dv = max(dv), .groups = "drop") |>
  dplyr::summarise(
    `published Cmax (mg/L)`     = 263.1,
    `simulated, no error`       = mean(cmax_ipred),
    `simulated, 24.5% prop err` = mean(cmax_dv)
  ) |>
  round(1)
```

  The published values are therefore reproduced from `Cc` (individual
  predictions) rather than from residual-error-perturbed observations.
* **Supplement Table S1 vs Table 2 on IIV for `Vmax`.** Supplementary Table 1
  describes model steps 6 and 7 as carrying "BSV CL, V1, V2, Vmax", but
  Section 3.2 and Table 2 both state that IIV on `Vmax` did not significantly
  improve the model and was not estimated. Table 2 governs: the model file has
  no `etalvmax`.
* **`V2` relative standard error.** Section 3.2 says 27%, Table 2 says 28%. This
  is cosmetic; RSEs are not encoded in the model file.

## Simulation assumptions

* **Virtual-cohort body weights** were drawn from a normal distribution with
  mean 65.5 kg and SD 18.1 kg truncated to 35-120 kg, per Na 2024 Section 2.6
  and Figure S1. The paper does not state the exact truncation rule; rejection
  sampling to the stated 35-120 kg range is used here.
* **Cohort size** is 200 per arm rather than the paper's 1000 virtual patients,
  per the nlmixr2lib vignette cap. The reported means are reproduced to within a
  few percent at this size.
* **Steady state** is taken at the dose given at 10080 h (60 weeks). The paper
  states steady state is reached in about 4-6 weeks; the longer run-in is used
  so that even the subjects in the upper tail of the 100.6 %CV `V2` distribution
  (whose peripheral redistribution is slow) are within 0.1% of steady state, as
  the check above asserts.
* **The second-dose 3-week interval** used in the actual trial (Na 2024
  Section 2.2) is not reproduced in the steady-state simulations, which use a
  uniform interval throughout, matching the paper's own simulation design.
* **Race and ethnicity** are not reported by the paper and are not modelled.
* **Dose group** was screened as a categorical covariate with an exponential
  form (Na 2024 Section 2.4) but was not retained in the final model, so it is
  not carried in `covariateData`.
* No parameter was tuned to improve agreement with any published value.
