---
title: "HOSU-53 (JBZ-001) DHODH inhibition in mouse and dog (Na 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HOSU-53 (JBZ-001) DHODH inhibition in mouse and dog (Na 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Na 2025 reports two independently fitted preclinical population PK/PD models for
the oral dihydroorotate dehydrogenase (DHODH) inhibitor HOSU-53 (JBZ-001) and its
plasma pharmacodynamic biomarker dihydroorotate (DHO) -- one in mice and one in
beagle dogs. Following the "replicate the author's structure" policy each is a
separate model file, and this single vignette walks the paper as a unit.

```{r models}
dog_mod   <- readModelDb("Na_2025_hosu53_dog")
mouse_mod <- readModelDb("Na_2025_hosu53_mouse")
dog_ui    <- rxode2::rxode(dog_mod)
mouse_ui  <- rxode2::rxode(mouse_mod)
```

* Citation: `r dog_ui$reference`
* Article: <https://doi.org/10.3390/pharmaceutics17040412>
* Dog model: `Na_2025_hosu53_dog` (Na 2025 Table 3)
* Mouse model: `Na_2025_hosu53_mouse` (Na 2025 Supplementary Table S3)

A rat dataset was collected but Na 2025 Section 3.3 explicitly excluded it from
the population modelling because dose proportionality was not met, so there is no
rat model file. Na 2025 also reports whole-body PBPK models built in PK-Sim
v11.0; those are not extracted here (see *Assumptions and deviations*).

# Population

The dog model was built from 507 HOSU-53 PK samples and 431 plasma DHO PD samples
in 39 male and female beagle dogs across four studies at 0.2, 0.3, 0.6, and
3 mg/kg PO (8, 12, 10, and 9 dogs respectively), with an initial single-dose study
at 10 mg/kg PO and 3 mg/kg IV. The 1 mg/kg PO once-daily arm of the 28-day GLP
study was excluded from the modelling dataset because of adverse gastrointestinal
findings (Na 2025 Section 3.1). 92 of the 431 DHO samples (21.3%), including
pre-dose samples, were below the limit of quantification and were excluded.

The mouse model was built from 309 PK samples and 253 PD samples in 45 mice
across three studies at 3 mg/kg IV and 4 and 10 mg/kg PO (9, 9, and 27 mice), a
mixture of non-tumour-bearing mice and triple-immunodeficient NCG mice carrying
MOLM-13 disseminated AML xenografts. Fewer than 4% of DHO samples were below the
limit of quantification.

Study-level detail is in Na 2025 Supplementary Table S1. Both models were fitted
with the SAEM algorithm in Monolix 2024R1 and evaluated by goodness-of-fit plots,
prediction-corrected VPCs, and a nonparametric bootstrap.

```{r population}
str(dog_ui$population, max.level = 1)
```

# Model structure

Both species share the same structure: two-compartment disposition with
first-order absorption, logit-normal bioavailability and linear elimination
(Na 2025 Section 2.5), coupled to an indirect-response turnover model for plasma
DHO in which HOSU-53 inhibits DHO degradation (Na 2025 Section 3.4, printed
equation):

$$\frac{dR}{dt} \;=\; k_{in} \;-\; k_{out}\cdot\left(1 - \frac{C_p^{\gamma}}{C_p^{\gamma} + IC_{50}^{\gamma}}\right)\cdot R,
\qquad k_{in} = R_0 \cdot k_{out}, \qquad R(0) = R_0$$

The printed final-model equation carries no $I_{max}$ term, i.e. $I_{max} = 1$;
the general form with $I_{max}$ appears only in Section 2.11 (the human
prediction). It is encoded as `limax <- fixed(log(1))` so the structure stays
explicit.

## Units

Na 2025 reports the volumes in mL and mL/h (Table 3, Table S3) and both
concentrations in umol/L (Table 3 units for `R0` and `IC50`; the Figure 1 and
Figure 2 axis labels). The model amount unit is therefore **nmol**, because
nmol / mL = umol/L. The paper does not state the dose-amount unit used in its
Monolix dataset and does not report the HOSU-53 molecular weight, so mg/kg doses
cannot be converted to model units from on-disk sources. Simulated doses below
are given in model amount units and anchored to a published observation (see
*Dog dose anchoring*).

```{r units}
dog_ui$units
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Na_2025_hosu53_dog.R` and
`inst/modeldb/specificDrugs/Na_2025_hosu53_mouse.R`. The table below collects
them in one place.

| Equation / parameter | Dog value | Mouse value | Source location |
|---|---|---|---|
| Two-compartment PK, first-order absorption, linear elimination | -- | -- | Section 3.3 ("two-compartment disposition models with first-order absorption and linear elimination") |
| DHO turnover with degradation inhibition (`d/dt(dho)`) | -- | -- | Section 3.4, printed equation (p. 10) |
| `logitfdepot` (F) | 0.67 | 0.53 | Table 3 / Table S3, row "F (%)"; logit scale per Section 2.5 |
| `lka` (Ka, 1/h) | 1.53 | 1.0 | Table 3 / Table S3, row "Ka (/h)" |
| `lcl` (CL, mL/h) | 150 (fixed) | 0.07 | Table 3 / Table S3, row "CL/F (mL/h)" |
| `lvc` (V1, mL) | 980 (fixed) | 1.2 | Table 3 / Table S3, row "V1 (mL)" |
| `lq` (Q, mL/h) | 400 (fixed) | 1.6 | Table 3 / Table S3, row "Q (mL/h)" |
| `lvp` (V2, mL) | 490 (fixed) | 1.2 | Table 3 / Table S3, row "V2 (mL)" |
| `lrbase` (R0, umol/L) | 0.06 | 0.01 | Table 3 / Table S3, row "R0 (umol/L)" |
| `lkout` (Kout, 1/h) | 52 | 155 | Table 3 / Table S3, row "Kout (/h)" |
| `lic50` (IC50, umol/L) | 0.1 | 1.55 | Table 3 / Table S3, row "IC50 (umol/L)" |
| `lhill` (gamma) | 1.9 | 1.71 | Table 3 / Table S3, row "gamma" |
| `limax` (Imax) | 1 (fixed) | 1 (fixed) | Section 3.4 equation carries no Imax term |
| `etalogitfdepot` | 0.24 (SD) | not estimated | Table 3, row "IIV F" |
| `etalka` | 0.59 (SD) | not estimated | Table 3, row "IIV Ka" |
| `etalcl` | 0.2 (SD, fixed) | 0.32 (SD) | Table 3 / Table S3, row "IIV CL" |
| `etalvc` | not estimated | 0.53 (SD) | Table S3, row "IIV V1" |
| `etalvp` | not estimated | 0.16 (SD) | Table S3, row "IIV V2" |
| `etalkout` | 0.42 (SD) | 0.62 (SD) | Table 3 / Table S3, row "IIV Kout" |
| `etalic50` | 0.44 (SD) | 0.53 (SD) | Table 3 / Table S3, row "IIV IC50" |
| `addSd` | 0.0033 (fixed) | not in model | Table 3, row "Additive residual error (PK)" |
| `propSd` | 0.45 (fixed) | 0.18 | Table 3 / Table S3, row "Proportional residual error (PK)" |
| `propSd_dho` | 0.55 | 0.46 | Table 3 / Table S3, row "Proportional residual error (PD)" |

Na 2025 was fitted in Monolix, whose results table reports `omega` on the
**standard-deviation** scale of the random effect -- the same scale as the
"proportional residual error" rows of the same tables, which are the SD-scale `b`
parameter of the Monolix error model. `nlmixr2`'s `ini()` takes the variance, so
each tabulated omega is squared in the model files (`etalka ~ 0.59^2`).

```{r ini-tables}
show_ini <- function(ui, species) {
  ui$iniDf |>
    dplyr::transmute(
      Species   = species,
      Parameter = name,
      Estimate  = ifelse(is.na(est), NA_real_, est),
      Fixed     = fix,
      Label     = ifelse(is.na(label), "", label)
    )
}
dplyr::bind_rows(show_ini(dog_ui, "dog"), show_ini(mouse_ui, "mouse")) |>
  knitr::kable(digits = 4, caption = "ini() entries of both packaged models.")
```

# Structural identity checks

These checks are exact consequences of the published equations, so they are
asserted rather than eyeballed. All simulations here are typical-value
(`omega = NA`, `sigma = NA`).

```{r helpers}
# Both models carry two residual-error endpoints (`Cc` and `dho`), and rxode2
# appends one pseudo-compartment per endpoint AFTER the ODE states, then
# requires every observation row to select one of them. The selector is `dvid`,
# not `cmt`: with `dvid` absent, an observation row naming an ODE state fails
# with "'dvid'->'cmt' or 'cmt' on observation record". Naming a single endpoint
# is sufficient -- rxSolve returns every output column (both `Cc` and `dho`) on
# those rows regardless of which endpoint `dvid` selects. So observation rows
# keep an ODE-state name in `cmt` and carry `dvid = 1L`; naming an algebraic
# observable in `cmt` instead would auto-inject a `cmt()` slot for it and
# renumber the ODE states, which is the failure mode this avoids.
add_obs <- function(ev, times) {
  rxode2::et(ev, times, cmt = "central")
}
with_dvid <- function(ev) {
  d <- as.data.frame(ev)
  d$dvid <- ifelse(d$evid == 0, 1L, NA_integer_)
  d
}
solve_typical <- function(mod, ev) {
  out <- rxode2::rxSolve(mod, with_dvid(ev), omega = NA, sigma = NA,
                         returnType = "data.frame")
  out[!duplicated(out$time), , drop = FALSE]
}
trap_auc <- function(time, conc) {
  sum(diff(time) * (utils::head(conc, -1) + utils::tail(conc, -1)) / 2)
}
```

## 1. Terminal half-life matches the analytic two-compartment solution

```{r check-thalf}
analytic_thalf <- function(cl, vc, q, vp) {
  k10 <- cl / vc; k12 <- q / vc; k21 <- q / vp
  s <- k10 + k12 + k21
  log(2) / (0.5 * (s - sqrt(s^2 - 4 * k21 * k10)))
}
thalf_dog_analytic   <- analytic_thalf(150, 980, 400, 490)
thalf_mouse_analytic <- analytic_thalf(0.07, 1.2, 1.6, 1.2)

sim_thalf <- function(mod, amt, tmax, tfit) {
  s <- solve_typical(mod, add_obs(rxode2::et(amt = amt, cmt = "central"),
                                  seq(0, tmax, by = tmax / 2000)))
  s <- s[s$time >= tfit & s$Cc > 0, ]
  unname(log(2) / -stats::coef(stats::lm(log(Cc) ~ time, data = s))[2])
}
thalf_dog_sim   <- sim_thalf(dog_mod,   20000, 96,  48)
thalf_mouse_sim <- sim_thalf(mouse_mod, 20,    336, 168)

stopifnot(
  abs(thalf_dog_sim   - thalf_dog_analytic)   < 0.01 * thalf_dog_analytic,
  abs(thalf_mouse_sim - thalf_mouse_analytic) < 0.01 * thalf_mouse_analytic
)
c(dog_analytic = thalf_dog_analytic, dog_simulated = thalf_dog_sim,
  mouse_analytic = thalf_mouse_analytic, mouse_simulated = thalf_mouse_sim)
```

## 2. Bioavailability is recovered exactly from the AUC ratio

`f(depot) <- fdepot` must make the oral-to-intravenous AUC ratio equal the
tabulated `F` for these linear models.

```{r check-f}
auc_route <- function(mod, cmtname, amt, tmax) {
  s <- solve_typical(mod, add_obs(rxode2::et(amt = amt, cmt = cmtname),
                                  seq(0, tmax, by = tmax / 4000)))
  trap_auc(s$time, s$Cc)
}
# Integrate over ~25 terminal half-lives per species. A much longer window
# lets the tail decay into solver noise and corrupts the trapezoid.
f_dog_recovered   <- auc_route(dog_mod,   "depot", 20000, 180) /
                     auc_route(dog_mod,   "central", 20000, 180)
f_mouse_recovered <- auc_route(mouse_mod, "depot", 20, 600) /
                     auc_route(mouse_mod, "central", 20, 600)

stopifnot(abs(f_dog_recovered   - 0.67) < 1e-3,
          abs(f_mouse_recovered - 0.53) < 1e-3)
c(dog = f_dog_recovered, mouse = f_mouse_recovered)
```

## 3. DHO holds at baseline with no drug on board

`kin = R0 * kout` and `dho(0) <- rbase` must make the undosed turnover system
sit exactly at `R0`.

```{r check-baseline}
hold <- function(mod) {
  s <- solve_typical(mod, add_obs(rxode2::et(), seq(0, 240, by = 1)))
  range(s$dho)
}
hold_dog   <- hold(dog_mod)
hold_mouse <- hold(mouse_mod)
stopifnot(all(abs(hold_dog - 0.06) < 1e-8), all(abs(hold_mouse - 0.01) < 1e-8))
rbind(dog = hold_dog, mouse = hold_mouse)
```

## 4. The DHO rise rate is capped by kin

Because the drug acts only on the *loss* term, the fastest DHO can rise -- even
under complete DHODH inhibition -- is `kin = R0 * kout`. This is what makes the
dog response markedly slower than its `kout` of 52/h would suggest.

```{r check-kin}
kin_dog   <- 0.06 * 52
kin_mouse <- 0.01 * 155
max_rise <- function(mod, amt) {
  s <- solve_typical(mod, add_obs(rxode2::et(amt = amt, cmt = "depot"),
                                  seq(0, 48, by = 0.01)))
  max(diff(s$dho) / diff(s$time))
}
# Deliberately supra-therapeutic amounts, so the inhibition term saturates and
# the rise rate is pushed against its kin ceiling.
rise_dog   <- max_rise(dog_mod,   240000)
rise_mouse <- max_rise(mouse_mod, 1000)
stopifnot(rise_dog <= kin_dog * 1.001, rise_mouse <= kin_mouse * 1.001)
data.frame(
  species  = c("dog", "mouse"),
  kin      = c(kin_dog, kin_mouse),
  max_rise = c(rise_dog, rise_mouse)
) |>
  knitr::kable(digits = 3, caption = "Maximum simulated dDHO/dt versus the kin ceiling (umol/L/h).")
```

# Dog dose anchoring

Model amounts are in nmol and the mg/kg-to-nmol conversion is unpublished (see
*Units*). To place the simulations on the scale the dogs were actually studied
at, the top dog arm is anchored to a **published observation**: Na 2025 Section 4
reports a maximum observed plasma DHO of about 5850 ng/mL in dogs. Converting
that to the model's umol/L needs the molecular weight of dihydroorotic acid,
158.11 g/mol -- a standard chemical constant for C5H6N2O4, **not** a value taken
from Na 2025 (see *Assumptions and deviations*) -- giving 37.0 umol/L. The
once-daily amount whose steady-state DHO Cmax equals that value is solved for
below and treated as the 3 mg/kg arm; the remaining arms are scaled by the
published dose ratios (0.2 : 0.3 : 0.6 : 3 mg/kg).

Two things this anchoring is **not**. It is not a molecular-weight conversion --
the mg/kg labels below identify which published dose *level* each arm
represents, and the nmol amount beside them is a simulation scale, not a
verified equivalent. And it assumes the paper's maximum observed dog DHO came
from the highest included dog arm (3 mg/kg once daily), which Na 2025 does not
state explicitly. Everything downstream that depends on absolute dose is
therefore conditional on that assumption; the half-life, bioavailability,
dose-proportionality, and baseline checks are not.

```{r anchor}
dho_cmax_ss <- function(amt, ndose = 6L) {
  ev <- add_obs(rxode2::et(amt = amt, cmt = "depot", ii = 24, addl = ndose - 1L),
                seq(24 * (ndose - 1L), 24 * ndose, by = 0.1))
  max(solve_typical(dog_mod, ev)$dho)
}
# 5850 ng/mL is Na 2025 Section 4. The 158.11 g/mol divisor is the standard
# molecular weight of dihydroorotic acid (C5H6N2O4), NOT a paper-reported value.
dho_target <- 5850 / 158.11   # umol/L
dog_top_amt <- stats::uniroot(
  function(a) dho_cmax_ss(a) - dho_target,
  interval = c(5000, 60000), tol = 50
)$root

dog_arms <- tibble::tibble(
  mgkg = c(0.2, 0.3, 0.6, 3),
  amt  = round(dog_top_amt * mgkg / 3)
) |>
  dplyr::mutate(treatment = sprintf("%.1f mg/kg level (%.0f nmol)", mgkg, amt))
dog_arms |>
  dplyr::rename("Published dose level (mg/kg)" = mgkg,
                "Model amount (nmol)" = amt,
                "Arm label" = treatment) |>
  knitr::kable(digits = 1, caption = "Anchored dog dose arms.")
```

# Virtual cohorts and simulation

Original observed data are not publicly available. No covariate was retained in
either final model (Na 2025 Section 3.3), so the virtual cohorts carry no
covariates -- the spread below is the models' published inter-individual
variability alone.

```{r cohort}
set.seed(20250325)
n_per_arm <- 60L

make_arms <- function(arms, ii, addl, obs_times) {
  out <- list()
  for (i in seq_len(nrow(arms))) {
    ev <- rxode2::et(amt = arms$amt[i], cmt = "depot", ii = ii, addl = addl) |>
      add_obs(obs_times) |>
      rxode2::et(id = seq_len(n_per_arm)) |>
      with_dvid()
    ev$id <- ev$id + (i - 1L) * n_per_arm
    ev$treatment <- arms$treatment[i]
    out[[i]] <- ev
  }
  dplyr::bind_rows(out)
}

# Coarse grid over the accumulation phase, fine grid over the last dosing
# interval (which is what Figures 1c / 2c plot).
dog_obs_times <- sort(unique(c(seq(0, 120, by = 2), seq(120, 144, by = 0.25))))
dog_events <- make_arms(dog_arms, ii = 24, addl = 5L, obs_times = dog_obs_times)
stopifnot(anyDuplicated(dog_events[, c("id", "time", "evid", "cmt")]) == 0L)

dog_sim <- rxode2::rxSolve(dog_mod, events = dog_events, keep = "treatment") |>
  as.data.frame() |>
  dplyr::distinct(id, time, .keep_all = TRUE)
stopifnot(!anyNA(dog_sim$Cc), !anyNA(dog_sim$dho))
```

```{r cohort-mouse}
set.seed(20250326)
mouse_arms <- tibble::tibble(
  amt = c(8, 20, 50),
  treatment = sprintf("%.0f nmol", c(8, 20, 50))
)
mouse_obs_times <- sort(unique(c(seq(0, 144, by = 4), seq(144, 168, by = 0.5))))
mouse_events <- make_arms(mouse_arms, ii = 24, addl = 6L, obs_times = mouse_obs_times)
stopifnot(anyDuplicated(mouse_events[, c("id", "time", "evid", "cmt")]) == 0L)
mouse_sim <- rxode2::rxSolve(mouse_mod, events = mouse_events, keep = "treatment") |>
  as.data.frame() |>
  dplyr::distinct(id, time, .keep_all = TRUE)
stopifnot(!anyNA(mouse_sim$Cc), !anyNA(mouse_sim$dho))
```

# Replicate published figures

Na 2025 Figure 1c and Figure 2c are prediction-corrected VPCs of dog HOSU-53 and
dog plasma DHO plotted against time after the last dose, both on a umol/L axis
spanning roughly 0.01-100 umol/L. The panels below plot the simulated median and
5th/95th percentiles over the same window; the reproduction is of the
concentration range and shape, not of the prediction correction itself (the
observed data behind the published pcVPC are not available).

```{r figure-1c-2c, fig.height = 6}
last_dose_t <- 24 * 5
dog_last <- dog_sim |>
  dplyr::filter(time >= last_dose_t, time <= last_dose_t + 24) |>
  dplyr::mutate(tald = time - last_dose_t) |>
  dplyr::select(id, treatment, tald, `HOSU-53` = Cc, DHO = dho) |>
  tidyr::pivot_longer(c(`HOSU-53`, DHO), names_to = "analyte", values_to = "conc") |>
  dplyr::group_by(analyte, treatment, tald) |>
  dplyr::summarise(
    Q05 = stats::quantile(conc, 0.05),
    Q50 = stats::quantile(conc, 0.50),
    Q95 = stats::quantile(conc, 0.95),
    .groups = "drop"
  )

ggplot(dog_last, aes(tald, Q50, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.18, colour = NA) +
  geom_line() +
  facet_wrap(~analyte, ncol = 1, scales = "free_y") +
  scale_y_log10() +
  scale_x_continuous(breaks = seq(0, 24, by = 4)) +
  labs(
    x = "Time (h) after last dose", y = "Plasma concentration (umol/L)",
    colour = "Dose arm", fill = "Dose arm",
    title = "Simulated dog HOSU-53 and plasma DHO at steady state",
    caption = "Comparable to Figure 1c (HOSU-53) and Figure 2c (DHO) of Na 2025."
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r figure-mouse, fig.height = 4}
mouse_sim |>
  dplyr::filter(time >= 144, time <= 168) |>
  dplyr::mutate(tald = time - 144) |>
  dplyr::group_by(treatment, tald) |>
  dplyr::summarise(
    Q05 = stats::quantile(Cc, 0.05), Q50 = stats::quantile(Cc, 0.50),
    Q95 = stats::quantile(Cc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(tald, Q50, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.18, colour = NA) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Time (h) after last dose", y = "HOSU-53 (umol/L)",
       colour = "Dose", fill = "Dose",
       title = "Simulated mouse HOSU-53 at steady state",
       caption = "Comparable to Supplementary Figure S1c of Na 2025.") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# PKNCA validation

Na 2025 Section 3.2 reports the non-compartmental results the models must be
consistent with: a mean elimination half-life of 6-8 h in dogs and 30 h in mice
after a single IV (3 mg/kg) and oral (10 mg/kg) dose, with mean bioavailability
of about 50% in dogs and 86% in mice. The NCA below is run on typical-value
single-dose profiles for each species and route.

```{r pknca}
single_dose <- function(mod, amt, tmax, route, species) {
  s <- solve_typical(mod, add_obs(
    rxode2::et(amt = amt, cmt = if (route == "IV") "central" else "depot"),
    seq(0, tmax, by = tmax / 400)))
  data.frame(id = 1L, time = s$time, Cc = s$Cc,
             treatment = paste(species, route), amt = amt)
}

nca_conc <- dplyr::bind_rows(
  single_dose(dog_mod,   20000, 100, "IV", "dog"),
  single_dose(dog_mod,   20000, 100, "PO", "dog"),
  single_dose(mouse_mod, 20,    336, "IV", "mouse"),
  single_dose(mouse_mod, 20,    336, "PO", "mouse")
) |>
  dplyr::filter(!is.na(Cc))

# Time-zero guarantee (extravascular pre-dose Cc = 0).
nca_conc <- dplyr::bind_rows(
  nca_conc,
  nca_conc |> dplyr::distinct(id, treatment, amt) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(treatment, id, time, .keep_all = TRUE) |>
  dplyr::arrange(treatment, id, time)

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(
  nca_conc |> dplyr::distinct(treatment, id, amt) |> dplyr::mutate(time = 0),
  amt ~ time | treatment + id
)
intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_wide <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "aucinf.obs", "half.life")) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
nca_wide |> knitr::kable(digits = 3, caption = "PKNCA results on the typical-value single-dose profiles.")
```

## Comparison against published NCA

```{r nca-compare}
published <- tibble::tribble(
  ~treatment,  ~half.life,
  "dog IV",    7.0,     # Section 3.2: "mean elimination half-life (T1/2) of 6 to 8 h" (rats and dogs)
  "dog PO",    7.0,
  "mouse IV", 30.0,     # Section 3.2: "a longer mean T1/2 of 30 h" (mice)
  "mouse PO", 30.0
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by        = "treatment",
  units     = c(half.life = "h"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Simulated vs. published half-life. * differs from reference by >20%.")
if (!is.null(attr(cmp, "footnote"))) attr(cmp, "footnote")
```

The dog model returns a terminal half-life of `r sprintf("%.1f", thalf_dog_sim)` h,
inside the 6-8 h range Na 2025 reports for dogs.

The mouse model returns `r sprintf("%.1f", thalf_mouse_sim)` h against the
reported mean NCA half-life of 30 h -- about 20% short, right at the flagging
tolerance. This is a property of the published parameters, not of the
extraction: `log(2)/beta` computed analytically from the Table S3 values of CL,
V1, Q, and V2 is `r sprintf("%.1f", thalf_mouse_analytic)` h, which the
simulation reproduces to four significant figures. The 30 h figure is a mean NCA
value from the single-dose study, whereas Table S3 is a simultaneous PK/PD SAEM
fit pooling three studies. Na 2025 itself flags the mouse half-life as the
anomaly in the dataset (Section 4: mouse elimination was much slower than in rats
and dogs, possibly through higher plasma protein binding, and "this did not fully
explain the PK differences in mice"). No parameter was adjusted.

## Bioavailability against the published NCA estimate

`F` is a structural parameter rather than an NCA parameter, so it is compared
separately.

```{r f-compare}
f_from_nca <- function(sp) {
  po <- nca_wide$aucinf.obs[nca_wide$treatment == paste(sp, "PO")]
  iv <- nca_wide$aucinf.obs[nca_wide$treatment == paste(sp, "IV")]
  stopifnot(length(po) == 1L, length(iv) == 1L)
  po / iv
}
tibble::tibble(
  Species                      = c("dog", "mouse"),
  `Model F (Table 3 / S3)`     = c(0.67, 0.53),
  `F recovered from simulated AUC` = c(f_from_nca("dog"), f_from_nca("mouse")),
  `Published NCA F (Section 3.2)`  = c(0.50, 0.86)
) |>
  knitr::kable(digits = 3, caption = "Bioavailability: model parameter, simulated AUC ratio, and the paper's own NCA estimate.")
```

The packaged models reproduce their own `F` exactly. The population estimates
differ from the paper's NCA-derived values in opposite directions (dog 0.67 vs
about 0.50; mouse 0.53 vs 0.86). This is a discrepancy **within** Na 2025, not an
extraction error: the NCA figures come from the single initial 10 mg/kg PO /
3 mg/kg IV study in 40% HPBCD, while the population estimates pool every study,
salt form, and vehicle. No parameter was tuned.

# Dose proportionality in dogs

Na 2025 Section 3.2 reports that dog exposure "increased in a dose-proportional
manner" over 0.3-3 mg/kg. Linear elimination makes this exact, which is a useful
regression guard on the model file.

```{r dose-prop}
dose_prop <- dog_arms |>
  dplyr::rowwise() |>
  dplyr::mutate(
    auc24_ss = {
      s <- solve_typical(dog_mod, add_obs(
        rxode2::et(amt = amt, cmt = "depot", ii = 24, addl = 5L),
        seq(120, 144, by = 0.1)))
      trap_auc(s$time, s$Cc)
    }
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(dose_normalised_auc = auc24_ss / amt)

stopifnot(diff(range(dose_prop$dose_normalised_auc)) <
            1e-6 * mean(dose_prop$dose_normalised_auc))
dose_prop |>
  dplyr::select(treatment, auc24_ss, dose_normalised_auc) |>
  dplyr::rename("Dose arm" = treatment,
                "AUC(0-24,ss) (umol*h/L)" = auc24_ss,
                "Dose-normalised AUC" = dose_normalised_auc) |>
  knitr::kable(digits = 6, caption = "Dog steady-state AUC is exactly dose-proportional.")
```

Na 2025 reports *less-than*-dose-proportional exposure in mice over 4-30 mg/kg
(Section 3.2), which a linear-elimination model cannot reproduce; the mouse model
was nevertheless fitted only over 3-10 mg/kg, where the linear assumption holds.

# DHO exposure and the toxicity threshold

The paper's central translational claim (Sections 3.2 and 4) is that severe
toxicity occurred in every species once the steady-state plasma DHO exposure
`AUCtau,ss` exceeded 1000 umol*h/L, that the optimal efficacy window is roughly
300-1000 umol*h/L, and that the highest DHO concentration observed **without**
toxicity in dogs was about 5850 ng/mL (37.0 umol/L). The dog model must place
that anchored exposure below the toxicity threshold.

```{r dho-exposure}
dho_exposure <- function(mod, amt, ndose = 6L) {
  s <- solve_typical(mod, add_obs(
    rxode2::et(amt = amt, cmt = "depot", ii = 24, addl = ndose - 1L),
    seq(24 * (ndose - 1L), 24 * ndose, by = 0.05)))
  data.frame(amt = amt, dho_cmax = max(s$dho),
             dho_auctau = trap_auc(s$time, s$dho))
}
dog_dho <- dplyr::bind_rows(lapply(dog_arms$amt, function(a) dho_exposure(dog_mod, a))) |>
  dplyr::mutate(treatment = dog_arms$treatment)

anchored <- dog_dho[nrow(dog_dho), ]
stopifnot(
  abs(anchored$dho_cmax - dho_target) < 0.02 * dho_target, # anchor solved correctly
  anchored$dho_auctau < 1000,                              # non-toxic per Section 4
  anchored$dho_auctau > 300                                # inside the efficacy window
)

dog_dho |>
  dplyr::select(treatment, dho_cmax, dho_auctau) |>
  dplyr::rename("Dose arm" = treatment,
                "DHO Cmax,ss (umol/L)" = dho_cmax,
                "DHO AUCtau,ss (umol*h/L)" = dho_auctau) |>
  knitr::kable(digits = 1,
               caption = "Simulated steady-state plasma DHO exposure in dogs. The toxicity threshold is 1000 umol*h/L and the efficacy window is 300-1000 umol*h/L (Na 2025 Sections 3.2 and 4).")
```

At the anchored top arm the model predicts a DHO `AUCtau,ss` of
`r sprintf("%.0f", anchored$dho_auctau)` umol*h/L -- inside the paper's stated
300-1000 umol*h/L therapeutic window and below the 1000 umol*h/L toxicity
threshold, consistent with Na 2025's report that no toxicity accompanied the
5850 ng/mL maximum DHO observation.

```{r dho-threshold}
threshold_amt <- stats::uniroot(
  function(a) dho_exposure(dog_mod, a)$dho_auctau - 1000,
  interval = c(dog_top_amt, 4 * dog_top_amt), tol = 50
)$root
c(anchored_top_dose_nmol = dog_top_amt,
  threshold_dose_nmol    = threshold_amt,
  fold_margin            = threshold_amt / dog_top_amt)
```

The 1000 umol*h/L threshold is crossed at only
`r sprintf("%.2f", threshold_amt / dog_top_amt)`-fold the anchored top arm. The
steepness comes from the `gamma` of 1.9 combined with the near-zero residual
degradation once `Cc` is far above the 0.1 umol/L `IC50`: DHO accumulation goes
from partially cleared to production-limited over a narrow dose band. That is
the same narrow margin Na 2025 describes clinically, where the dog NOAEL of
0.6 mg/kg once daily sits one dose level below the adverse 1 mg/kg once-daily
arm.

## Species comparison of the PD parameters

```{r pd-species}
tibble::tibble(
  Parameter = c("R0 (umol/L)", "Kout (1/h)", "kin = R0*Kout (umol/L/h)",
                "IC50 (umol/L)", "gamma"),
  Dog   = c(0.06, 52, 0.06 * 52, 0.10, 1.90),
  Mouse = c(0.01, 155, 0.01 * 155, 1.55, 1.71)
) |>
  dplyr::mutate(`Mouse / dog` = Mouse / Dog) |>
  knitr::kable(digits = 3, caption = "DHO turnover parameters by species (Na 2025 Table 3 and Table S3).")
```

The Discussion states that the mouse and dog PD parameters "showed overall
comparability". On the published numbers only the sigmoidicity `gamma` is close
(1.71 vs 1.90); the baseline `R0` differs 6-fold, `Kout` 3-fold, and `IC50`
15-fold. The values above are transcribed verbatim from Table 3 and
Supplementary Table S3 and were re-checked against the supplement PDF.

# Assumptions and deviations

- **Dose-amount units are not published.** Na 2025 reports volumes in mL and
  concentrations in umol/L but never states the amount unit of its Monolix
  dataset, and does not report the HOSU-53 molecular weight. The model files
  declare `units$dosing = "nmol"` because nmol / mL = umol/L is the only
  dimensionally consistent reading of the published tables. Converting a mg/kg
  dose into model units requires the unpublished molecular weight, so the dog
  arms in this vignette are anchored to the paper's reported maximum observed
  dog DHO concentration (5850 ng/mL = 37.0 umol/L, Section 4) rather than stated
  in mg/kg. That anchoring additionally assumes the maximum observed DHO came
  from the highest included dog arm (3 mg/kg once daily), which Na 2025 does not
  state; the mg/kg labels on the simulated arms identify the published dose
  *level* each represents and are not a molecular-weight conversion. The mouse
  arms are given in model units only. Nothing in the model files depends on this
  anchoring -- it affects only the illustrative dose scale in this vignette.
- **One non-paper constant is used, and only in this vignette.** The 5850 ng/mL
  to 37.0 umol/L conversion above divides by 158.11 g/mol, the standard
  molecular weight of dihydroorotic acid (C5H6N2O4). That constant is *not*
  reported in Na 2025; it is standard chemistry, introduced here solely to put
  the paper's ng/mL observation on the models' umol/L scale. No parameter in
  either model file depends on it. Na 2025 does contain one internal ng/mL to
  micromolar pairing -- Section 4's "~25 uM DHO (or ~3200 ng/mL)", quoted for a
  different compound's mouse study -- which implies about 128 g/mol and so
  disagrees with the true molecular weight by roughly 24%; both numbers in that
  parenthetical are tilde-prefixed approximations, so it is read as the paper
  rounding loosely rather than as a competing value. The choice does not change
  any conclusion drawn here: re-anchoring the dog arms with 128 g/mol instead of
  158.11 raises the top arm's DHO `AUCtau,ss` from 626 to 826 umol*h/L, which is
  still inside the paper's 300-1000 umol*h/L window and still below the
  1000 umol*h/L toxicity threshold.
- **`dho` and `DHO` were registered as new canonical compartment names.** The
  plasma-dihydroorotate turnover state had no canonical name. Rather than
  declaring it per-paper, `dho` (the lowercase ODE state) and `DHO` (the
  uppercase observation-output sibling) were added to
  `inst/references/compartment-names.md` under *Endogenous metabolic species*,
  with these two model files as the founding examples, so that later
  DHODH-inhibitor extractions reuse the name. The lowercase-state /
  uppercase-observable split follows the registered `ANC` / `anc` and
  `PRU` / `pru` pairs and is required because `checkModelConventions()` enforces
  lowercase ODE-state names. These models observe the lowercase state directly
  (`dho ~ prop(propSd_dho)`), which is equally permitted.
- **The mouse half-life runs about 20% short of the published NCA value.**
  Analytically, `log(2)/beta` from the Table S3 disposition parameters is 24.0 h
  against the 30 h mean NCA half-life quoted in Section 3.2. The discrepancy is
  in the published numbers (a pooled three-study SAEM fit versus a single-study
  mean NCA), not in the transcription, and Na 2025 Section 4 already identifies
  mouse PK as the unexplained part of its dataset. No parameter was tuned.
- **IIV is on the standard-deviation scale.** Na 2025 was fitted in Monolix,
  whose results table reports `omega` as the standard deviation of the random
  effect -- the same scale as the "proportional residual error" rows of Table 3
  and Table S3, which are the SD-scale `b` parameter of the Monolix error model.
  The paper does not say so explicitly. `nlmixr2` takes the variance, so each
  tabulated value is squared in `ini()`.
- **`Imax` is fixed at 1.** The final-model equation printed in Section 3.4
  carries no `Imax` term. The general form including `Imax` appears only in
  Section 2.11 (human PD prediction). `limax <- fixed(log(1))` records this
  explicitly rather than dropping the term.
- **`CL/F` is encoded as `CL`.** Table 3 and Table S3 label the clearance row
  "CL/F", but both tables also report `F` as a separate estimated parameter, and
  both fits used pooled IV and PO data (the dog disposition parameters came from
  an IV-only fit). The value is therefore a true clearance. The dimensional
  check supports this: the resulting terminal half-lives are 7.1 h (dog) and
  24.0 h (mouse), matching the 6-8 h and 30 h NCA values in Section 3.2. Reading
  it as `CL/F` instead would shorten both half-lives by a factor of `F`.
- **`Q` is in mL/h, not L/h.** The `Q` row header of Table 3 and Table S3 reads
  "Q (mL/h)", but the abbreviation footnote of both tables reads "Q,
  intercompartment clearance (L/h)". mL/h is the correct reading: it is the unit
  consistent with `CL` and `V1` in the same tables, 400 L/h is physiologically
  impossible in a dog, and only mL/h reproduces the published NCA half-lives.
- **No erratum applies.** PubMed (PMID 40284407) and Europe PMC list no
  correction, corrigendum, or erratum for this article as of the extraction date.
- **No covariates.** Sex, dose level, salt form, and vehicle formulation were
  screened (Section 2.7) and none were retained (Section 3.3). They are recorded
  in each model file's `covariatesDataExcluded` metadata so the screen's
  provenance survives, but they are absent from `model()`.
- **The rat model does not exist.** Na 2025 Section 3.3 excluded the rat data
  from the population modelling because dose proportionality was not met.
- **The PK-Sim PBPK models are not extracted.** Na 2025 built whole-body PBPK
  models for rat, dog, and human in PK-Sim v11.0, but the only model inputs
  published are the compound-specific properties in Table 1 and Supplementary
  Table S2 (lipophilicity, fu, pKa, solubility, intestinal permeability,
  specific clearance, blood:plasma ratio). The physiological structure,
  organ volumes, blood flows, partition coefficients, and cellular
  permeabilities are supplied by the PK-Sim platform and are not written out in
  any on-disk source, so the PBPK models are not reproducible here without
  substituting platform defaults. Table 4 (time above the 669 pM hepatic target)
  and Figure 3 (human plasma and hepatic profiles) are PBPK outputs and are
  likewise not reproduced.
- **The human PD prediction is not reproduced.** Section 2.11 predicts human DHO
  by driving the dog PD parameters with the PBPK-predicted human HOSU-53 plasma
  profile. That driving profile is a PBPK output (see above), so the quoted human
  results -- DHO Cmax of about 791 ng/mL at 5 mg and 5500 ng/mL at 25 mg, and an
  `AUCtau,ss` of 623 umol*h/L at 25 mg once daily -- cannot be regenerated from
  the packaged models alone.
- **The published pcVPCs are not prediction-corrected here.** Figures 1c and 2c
  of Na 2025 are prediction-corrected VPCs against observed data that are not
  publicly available. The figures above show the simulated median and 5th/95th
  percentiles over the same time-after-last-dose window, which reproduces the
  concentration range and shape but not the correction.
- **Multi-endpoint event tables select the endpoint with `dvid`.** Both models
  carry two residual-error endpoints (`Cc` and `dho`), and rxode2 appends one
  pseudo-compartment per endpoint after the ODE states, then requires every
  observation row to select one of them. The selector is the `dvid` column, not
  `cmt`: with `dvid` absent, an observation row naming an ODE state fails with
  `'dvid'->'cmt' or 'cmt' on observation record`. Every event table here
  therefore keeps ODE-state names in `cmt` (`depot` / `central`) on both dose
  and observation rows and adds `dvid = 1L` on the observation rows. Selecting
  one endpoint is sufficient -- rxSolve returns both `Cc` and `dho` on those
  rows. Putting an algebraic observable name in `cmt` also runs, but it
  auto-injects a compartment slot for that observable and renumbers the ODE
  states, so it is avoided here.
- **Baseline DHO relative to the assay.** The dog `R0` of 0.06 umol/L is
  essentially the dog DHO assay LLOQ of 10 ng/mL (0.063 umol/L), consistent with
  Section 3.1's report that 21.3% of dog DHO samples -- including pre-dose
  samples -- were below quantification. The mouse `R0` of 0.01 umol/L
  (1.6 ng/mL) sits below the 10 ng/mL mouse assay LLOQ and is therefore an
  extrapolated baseline identified mainly by post-dose data. Both are used as
  published.
