---
title: "NA_NA_tte_lognormal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NA_NA_tte_lognormal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(ggplot2)
```

# Model and source

* Citation: `r readModelDb("NA_NA_tte_lognormal")()$reference`
* Description: `r readModelDb("NA_NA_tte_lognormal")()$description`
* Source: BAST Inc Limited, "BAST approach to parametric time-to-event (PTTE) modelling," internal guiding document, 12 July 2017 (`BAST_PTTE_modelling.pdf` shipped in the DDMORE bundle).
* DDMORE Foundation Model Repository entry: [DDMODEL00000243](https://repository.ddmore.eu/model/DDMODEL00000243)
* Source bundle (local mirror): `dpastoor/ddmore_scraping/243/`
* Linked publication: **none.** The bundle is a methodological teaching example built on entirely simulated data; the BAST guiding-document text states "there is not yet a publication to go along with the model" (`Model_Accommodations.txt`).

This vignette validates the BAST 2017 PTTE Competing Event 1 hazard
model packaged under `inst/modeldb/ddmore/NA_NA_tte_lognormal.R`
(NONMEM run name `runCOMPEV1_101`). Competing Event 1 in the bundle is
interval-censored; the BAST guiding document Section  2.4.1 (Figure 2-3)
selected log-normal as the base distribution by AIC, and Section  2.4.2
(Table 2-4) retained AGE as the only covariate.

For sibling Event 1, Event 2, and Competing Event 2 hazard models from
the same bundle, see `NA_NA_tte_gompertz.R`,
`NA_NA_tte_gompertz_ev2.R`, and `NA_NA_tte_loglogistic.R`.

# Population

The BAST 2017 PTTE bundle is a methodological teaching example with
N = 200 simulated patients, four timed event types, and six baseline
covariates (BAST Section  2.2.2). Of the 200 simulated patients, 36 (18%)
experienced Competing Event 1 (BAST Table 2-1) -- the lowest event
incidence of the four event types. Competing Event 1 is **interval-
censored**: exact event times are unknown, only that the event
occurred between two scheduled assessment visits.

```{r population-meta}
m <- readModelDb("NA_NA_tte_lognormal")
str(m()$meta$population, max.level = 1)
```

# Source trace

| Equation / parameter             | Value      | Source location                                                                  |
|----------------------------------|------------|----------------------------------------------------------------------------------|
| Hazard form `h(t) = val * pdf(t) / (1 - Phi(num))` | n/a | Executable_runCOMPEV1_101.mod $PK / $DES (`pdf = (1/(fac*lambda*(T+DEL)))*exp(-(num**2)/(2*lambda*lambda))`, `DADT(1) = VAL*pdf/(1-phi(num/lambda))`); BAST guiding doc Section  2.4.1 confirms log-normal selected |
| `llambda_compev1` (sigma)        | log(1.42)  | Output_simulated_runCOMPEV1_101.res FINAL TH1 = 1.42E+00; not rescaled |
| `lalpha_compev1` (median scale)  | log(984)   | Output_simulated_runCOMPEV1_101.res FINAL TH2 = 9.84E+02; not rescaled |
| `e_age_compev1` (AGE coefficient)| 50.9       | Output_simulated_runCOMPEV1_101.res FINAL TH3 = 5.09E+01; rescaled by /1000, applied to `(AGE - 55)` |
| eta on `lambda` (`OMEGA(1,1)`)     | 0 FIXED    | Output_simulated_runCOMPEV1_101.res FINAL OMEGA(1,1) = 0.00E+00 (placeholder; no estimated IIV) |
| Covariate-selection DeltaOFV         | -14.822    | BAST guiding doc Section  2.4.2, Table 2-4 (AGE chosen as final covariate model COMPEV1_101) |

# Virtual cohort

```{r cohort}
set.seed(20260506)

n_subjects <- 50
cohort_subjects <- tibble(
  id  = seq_len(n_subjects),
  AGE = pmin(pmax(round(rnorm(n_subjects, mean = 58.7, sd = 12)), 24), 84)
)

obs_grid <- tibble(
  time = c(0, seq(7, 800, by = 14)),
  evid = 0L,
  amt  = 0
)

events <- tidyr::crossing(cohort_subjects, obs_grid)
events <- events[, c("id", "time", "evid", "amt", "AGE")]

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
cat("Cohort: ", n_subjects, " subjects, ", nrow(events), " event rows\n", sep = "")
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(m, events = events) |>
  as.data.frame()
```

# Replicate published behaviour -- typical-value survival trajectory

The BAST guiding document Section  2.4.1 (Figure 2-3) reports a Kaplan-Meier
curve of Competing Event 1 vs. the candidate parametric distributions;
log-normal was chosen (AIC -2.144 vs. exponential 0). The covariate-
VPC stratifications by patient age (Figure 2-11, Figure 2-12) cover
the window 0 -- 400 days. Competing Event 1 has the slowest event rate
of the four events; at the typical AGE (55 years), the model's
typical-value survival drops from `S(0) = 1` to about `S(800) ~= 0.57`.

```{r km-style}
sim |>
  group_by(time) |>
  summarise(
    median_sur = median(sur),
    q05        = quantile(sur, 0.05),
    q95        = quantile(sur, 0.95),
    .groups    = "drop"
  ) |>
  ggplot(aes(time, median_sur)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue") +
  labs(
    x        = "Time (days)",
    y        = "Survival probability",
    title    = "Competing Event 1 typical-value survival (median + 5-95% range over the virtual cohort)",
    caption  = "Compare against BAST 2017 PTTE guiding-document Figure 2-3 / Figure 2-11."
  ) +
  scale_y_continuous(limits = c(0, 1))
```

# Mechanistic sanity checks (verification-checklist Section  F.3)

## F.3.1 -- Hazard has the characteristic log-normal shape (rises then falls)

The log-normal hazard rises from 0 at `t = 0`, peaks near `t = alpha`
(the median time scale), and decays at long times. The check below
confirms this characteristic non-monotonic shape for a typical subject.

```{r hazard-shape}
ev_typ <- tibble(id = 1L, time = c(0, seq(7, 1500, by = 7)),
                 evid = 0L, amt = 0, AGE = 55)
sim_typ <- rxode2::rxSolve(m, events = ev_typ) |> as.data.frame()
ggplot(sim_typ, aes(time, hazard)) +
  geom_line(colour = "firebrick") +
  geom_vline(xintercept = 984, colour = "grey60", linetype = "dotted") +
  labs(x = "Time (days)", y = "Instantaneous hazard (1/day)",
       title = "Competing Event 1 typical-subject hazard trajectory",
       caption = "Log-normal hazard: rises, peaks near alpha = 984 days, decays.")

# The hazard at the peak should be larger than at both early (t = 50)
# and late (t = 1500) times.
hazard_peak  <- max(sim_typ$hazard)
hazard_early <- sim_typ$hazard[sim_typ$time == 50]
hazard_late  <- sim_typ$hazard[sim_typ$time == 1500]
stopifnot(hazard_peak > hazard_early)
stopifnot(hazard_peak > hazard_late)
```

## F.3.2 -- Older patients have higher hazard (positive AGE coefficient)

The BAST guiding document Section  2.4.2 (Table 2-4) reports a +50.9
coefficient on the centred `(AGE - 55)/1000` term, meaning older
patients have a higher hazard. The check below confirms the simulated
trajectories shift in the expected direction.

```{r covariate-shift}
make_alt <- function(label, age) {
  tibble(id = match(label, c("AGE = 35", "AGE = 55", "AGE = 75")),
         time = c(0, seq(7, 800, by = 14)),
         evid = 0L, amt = 0, AGE = age, scenario = label)
}
scenarios <- bind_rows(
  make_alt("AGE = 35", 35),
  make_alt("AGE = 55", 55),
  make_alt("AGE = 75", 75)
)
sim_scen <- rxode2::rxSolve(m, events = scenarios, keep = c("scenario")) |>
  as.data.frame()

ggplot(sim_scen, aes(time, sur, colour = scenario)) +
  geom_line() +
  labs(x = "Time (days)", y = "Survival probability",
       title = "Competing Event 1 typical-subject survival under one-at-a-time AGE shifts",
       colour = "Scenario") +
  scale_y_continuous(limits = c(0, 1))

final_sur <- sim_scen |>
  filter(time == max(time)) |>
  select(scenario, sur)
print(final_sur)

young_sur <- final_sur$sur[final_sur$scenario == "AGE = 35"]
ref_sur   <- final_sur$sur[final_sur$scenario == "AGE = 55"]
old_sur   <- final_sur$sur[final_sur$scenario == "AGE = 75"]

# Older patients -> higher hazard -> lower survival
stopifnot(old_sur < ref_sur)
stopifnot(young_sur > ref_sur)
```

## F.3.3 -- Final-fit objective-function value matches the bundle

`Output_simulated_runCOMPEV1_101.res` reports `OBJV = 360253.678` at
the final estimates. This value is informational here.

# Self-consistency with the bundle's simulated dataset (F.2)

A full F.2 self-consistency check would re-simulate the bundle's
shipped `Simulated_event_data.csv` (200 subjects, DVID = 3 records)
under the nlmixr2lib model and compare against the bundle's
`Output_simulated_runCOMPEV1_101.res` `$TABLE` output. The bundle
dataset is outside this package and not redistributed.

# Assumptions and deviations

* **Numerical rescalings preserved.** The .mod uses an internal /1000
  rescaling on the AGE coefficient only; lambda and alpha enter the
  hazard at their raw THETA values. The biologically meaningful values
  are:
  - Log-normal lambda (sigma): 1.42 (unitless)
  - Log-normal alpha (median time scale): 984 days
  - AGE coefficient: 50.9 / 1000 = 0.0509 per year above 55

* **`pnorm()` used in place of NONMEM `phi()`.** The .mod $DES uses
  NONMEM's intrinsic `phi()` for the standard normal CDF. rxode2's
  equivalent is `pnorm()`; the two are mathematically identical.

* **Small-time offset DEL = 1e-8 preserved.** The .mod $DES adds a
  small offset to t inside `log((t+DEL)/alpha)` to keep the expression
  finite at `t = 0`. We preserve the offset.

* **No estimated IIV.** The source `$OMEGA` is `0 FIX`. No inter-
  subject variability is in the BAST PTTE 2017 guiding-document model.

* **Interval-censored event times.** Competing Event 1 events are
  observed only between scheduled assessment visits. The packaged
  nlmixr2lib model is intended for forward simulation; users who want
  to honour the interval-censoring convention should snap simulated
  event times to the next assessment visit (BAST guiding doc Section  2.3.3).

* **No publication-PDF cross-check.** The BAST PTTE 2017 guiding
  document is the only source of equations and methodological
  narrative; no peer-reviewed publication exists.

* **Convention warning.** `nlmixr2lib::checkModelConventions()` flags
  the `units$concentration` field (the TTE output `sur` is a survival
  probability, not a mass/volume concentration). The compartment is
  named `cumhaz` (canonical TTE auxiliary-state name).
