---
title: "NA_NA_sunitinib"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NA_NA_sunitinib}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* DDMORE Foundation Model Repository entry:
  [DDMODEL00000231](https://repository.ddmore.eu/model/DDMODEL00000231)
  (MPD6: Sutent / sunitinib semi-mechanistic PK/PD in non-small cell
  lung cancer; MDL/PharmML deposit, version 3 in the dpastoor scrape).
* Linked publication: not located on disk. The bundle ships only
  `Sunitinib_MPD6_model.mdl` (the MDL source) and
  `Sunitinib_MPD6_model.xml` (the auto-rendered PharmML 0.6.1).
  No `Output_real_*.lst`, no `Output_simulated_*.lst`, no
  `Simulated_*.csv`, and no `Model_Accomodations.text`.
* Deposit description (`DDMODEL00000231.rdf`,
  `model-has-description`): "This is a semi-mechanistic PK/PD model for
  sunitinib therapy in non-small cell lung cancer patients. It was
  developed and validated using clinical trial data provided by the
  drug developer."

This vignette is the validation companion to
`inst/modeldb/ddmore/NA_NA_sunitinib.R`. Because the bundle ships no
NONMEM listing, no companion paper, and no simulated reference dataset,
the validation strategy reduces to **F.2 self-consistency** (model
parses, `rxode2::rxSolve()` runs to completion at typical-value
parameters, the trajectory is dynamically plausible). No side-by-side
comparison against published numerical or graphical values is
performed. See the Errata section.

# Population

The DDMORE bundle does not expose subject counts, study counts,
age/weight/sex/race distributions, or dose-level breakdowns. The
deposit's RDF metadata describes the modelled disease as non-small cell
lung cancer (NSCLC); the standard sunitinib oncology regimen at the
time of MPD6 deposit was 50 mg PO QD on a 4-weeks-on / 2-weeks-off
schedule (gastrointestinal stromal tumour and metastatic renal cell
carcinoma labels) or 37.5 mg PO QD continuously (later label
expansion); the MDL itself is dose-input-agnostic.

The same information is available programmatically via the model's
`population` metadata
(`readModelDb("NA_NA_sunitinib")$population` after
`devtools::load_all()` against the worktree).

# Source trace

All parameter values come from `Sunitinib_MPD6_model.mdl`'s `parObj`
`STRUCTURAL{}` and `VARIABILITY{}` blocks. Per the operator decision
recorded in the queue sidecar (`response-001.json`, value
`extract_mdl`), these are treated as the deposited final estimates.
The `.mod`/`.lst` cross-check that the DDMORE-source extraction skill
normally requires is unavailable for this bundle; see Errata.

| Equation / parameter | Value | Source location (DDMODEL00000231) |
|----------------------|-------|------------------------------------|
| Structural model: 15-state ODE (parent + metabolite 2-cmt PK with separate depots; 4 indirect-response biomarkers; tumor-volume sphere with capped exponential growth; resistance, lat-signal, parent-integrator, lambda-feedback memory chains) | n/a | `.mdl` `MPD6_ODE_mdl` `MODEL_PREDICTION{}` `DEQ{}` block (A1..A15) |
| Two-depot dosing with effective bioavailability split | n/a | `.mdl` `DEQ{}` `A14: deriv = -Kah*A14, init = D` and `A15: deriv = -Kamh*A15, init = D`; A1 receives `Kah*A14*(1-fp)`, A3 receives `Kamh*A15*fp` |
| Hard-coded `fp = 0.21` | 0.21 | `.mdl` `MODEL_PREDICTION{}` `fp=0.21` |
| Hard-coded `th1..4 = 1` (parent-only drive on biomarkers) | 1.0 | `.mdl` `MODEL_PREDICTION{}` `th1=1`..`th4=1` |
| Hard-coded `thettum = 1` (parent-only drive on tumor) | 1.0 | `.mdl` `MODEL_PREDICTION{}` `thettum=1` |
| Hard-coded `dres = 0` (no resistance decay) | 0 | `.mdl` `MODEL_PREDICTION{}` `dres=0` |
| Tumor doubling-time cap `maxgr = ln(2) / 30` | n/a | `.mdl` `MODEL_PREDICTION{}` `maxgr=ln(2)/30` |
| Tumor-state-to-radius observation `((3/(4*pi))*max(A9,0))^(1/3)` | n/a | `.mdl` `output7 = ((3/(4*3.1416))*max(A9,0))^(1/3)` |
| `lka = log(0.0715)` | 0.0715 | `.mdl` parObj `POP_Ka` |
| `lcl = log(26.9)` | 26.9 | `.mdl` parObj `POP_Cl` |
| `lvc = log(3220)` | 3220 (= 3.22e3) | `.mdl` parObj `POP_V1` |
| `lq = log(17.5)` | 17.5 | `.mdl` parObj `POP_QQ` |
| `lvp = log(127)` | 127 | `.mdl` parObj `POP_V2` |
| `lka_metab = log(0.177)` | 0.177 | `.mdl` parObj `POP_Kam` |
| `lcl_metab = log(15.6)` | 15.6 | `.mdl` parObj `POP_Clm` |
| `lvc_metab = log(3710)` | 3710 (= 3.71e3) | `.mdl` parObj `POP_Vm1` |
| `lq_metab = log(159)` | 159 | `.mdl` parObj `POP_QQm` |
| `lvp_metab = log(156)` | 156 | `.mdl` parObj `POP_Vm2` |
| `lkout_biom1..4 = log(0.111, 0.101, 0.169, 0.00659)` | 0.111 / 0.101 / 0.169 / 0.00659 | `.mdl` parObj `POP_d1`..`POP_d4` |
| `lpd1..4 = log(144, 22, 36.4, 98.1)` | 144 / 22 / 36.4 / 98.1 | `.mdl` parObj `POP_pd1`..`POP_pd4` |
| `lst1..4 = 0` (baseline = 1) | 0 | `.mdl` parObj `POP_st1`..`POP_st4` |
| `lx0 = 0` (initial radius = 1) | 0 | `.mdl` parObj `POP_x0` |
| `lpdm = log(0.129)` | 0.129 | `.mdl` parObj `POP_pdm` |
| `llam = log(0.000193)` | 0.000193 | `.mdl` parObj `POP_lam` |
| `llam0 = log(0.00189)` | 0.00189 | `.mdl` parObj `POP_lam0` |
| `lalphres = log(0.0102)` | 0.0102 | `.mdl` parObj `POP_alphres` |
| `lpdr = log(0.0286)` | 0.0286 | `.mdl` parObj `POP_pdr` |
| IIV variance for non-zero etas (var = sd^2 from MDL `type is sd`) | (see model file) | `.mdl` parObj VARIABILITY block |
| Residual error: `propSd` (b_1), `propSd_Cc_metab` (b_2), `propSd_biom1..4` (b_3..b_6), `addSd_tumorRadius` (a_7), `propSd_tumorRadius` (b_7) | 0.512 / 0.429 / 0.503 / 0.137 / 0.28 / 0.135 / 0.241 / 0.0856 | `.mdl` parObj `b_1`..`b_6`, `a_7`, `b_7` |

The `.mdl`'s 14×14 `OMEGA` `type is corr` correlation matrix carries
small (mostly |0.01|–|0.07|) off-diagonal entries; those off-diagonal
correlations are not encoded in the nlmixr2 port (see Errata —
"Off-diagonal IIV correlations dropped").

# Self-consistency simulation (F.2 substitute)

The DDMORE bundle ships no simulated dataset, so the canonical F.2
recipe ("re-simulate the bundle's `Simulated_*.csv` and visually
compare with `Output_simulated_*.lst`") cannot be run. The substitute
below verifies that the model parses, `rxode2::rxSolve()` runs to
completion at typical-value parameters, and the trajectory is
dynamically plausible (parent and metabolite plasma concentrations
rise after dose and decay; biomarker 1 rises in response to drug
exposure via Kout inhibition; biomarkers 2–4 fall via Kin inhibition;
tumor volume grows slowly).

```{r typical-cohort}
mod  <- readModelDb("NA_NA_sunitinib")
modT <- rxode2::zeroRe(mod)

# Single-subject, typical-value 28-day simulation. The MDL dosing
# convention is two parallel oral depots (parent and metabolite) that
# both receive the dose D at every dosing event; the parent depot
# absorbs into central with effective bioavailability (1 - fp) = 0.79
# and the metabolite depot absorbs into central_metab with effective
# bioavailability fp = 0.21 (fp hard-coded in the MDL, preserved in
# the nlmixr2 port). Sunitinib's standard daily dose at MPD6 deposit
# was 50 mg PO QD; one daily dose is shown here. Time unit is day.
times <- seq(0, 28, by = 0.25)
n_obs <- length(times)
events <- data.frame(
  id   = 1L,
  time = c(0, 0, times),
  evid = c(1L, 1L, rep(0L, n_obs)),
  amt  = c(50, 50, rep(NA_real_, n_obs)),
  cmt  = c("depot", "depot_metab", rep("Cc", n_obs))
)

sim <- rxode2::rxSolve(modT, events = events) |> as.data.frame()
```

```{r typical-pk-plot, fig.width = 7, fig.height = 4}
pk <- sim |>
  dplyr::select(time, Cc, Cc_metab) |>
  tidyr::pivot_longer(c(Cc, Cc_metab),
                      names_to = "analyte", values_to = "conc")

ggplot(pk, aes(time, conc, colour = analyte)) +
  geom_line() +
  labs(x = "Time (days)",
       y = "Concentration (model-internal units)",
       title = "Parent + metabolite plasma concentrations after one 50 mg dose",
       caption = paste("Two-depot oral dose with hard-coded fp = 0.21",
                       "metabolite split. Concentration units are bundle-internal",
                       "(volumes are in mL); the absolute scale is unverified",
                       "(no .lst, no publication).")) +
  theme_minimal()
```

```{r typical-biom-plot, fig.width = 7, fig.height = 5}
biom <- sim |>
  dplyr::select(time, biom1, biom2, biom3, biom4) |>
  tidyr::pivot_longer(c(biom1, biom2, biom3, biom4),
                      names_to = "biomarker", values_to = "conc")

ggplot(biom, aes(time, conc)) +
  geom_line() +
  facet_wrap(~biomarker, scales = "free_y") +
  labs(x = "Time (days)",
       y = "Biomarker level (relative to baseline = 1)",
       title = "Indirect-response PD biomarker trajectories",
       caption = paste("biom1 is Kout-modulated (drug INCREASES Kout-inhibition;",
                       "biomarker rises). biom2-4 are Kin-modulated (drug DECREASES",
                       "Kin; biomarkers fall). The MDL does not name the four",
                       "biomarkers; identity is unconfirmed.")) +
  theme_minimal()
```

```{r typical-tumor-plot, fig.width = 7, fig.height = 4}
tum <- sim |>
  dplyr::select(time, tumor, tumorRadius) |>
  tidyr::pivot_longer(c(tumor, tumorRadius),
                      names_to = "state", values_to = "value")

ggplot(tum, aes(time, value)) +
  geom_line() +
  facet_wrap(~state, scales = "free_y") +
  labs(x = "Time (days)",
       y = "Tumor state",
       title = "Tumor volume and radius trajectories",
       caption = paste("Tumor state is volume (mm^3 or model-internal units);",
                       "tumorRadius is the observation Y7 = (3/(4*pi)*max(tumor,0))^(1/3).")) +
  theme_minimal()
```

```{r f2-sanity-checks}
sim_at <- function(t) sim[which.min(abs(sim$time - t)), ]

# Parent + metabolite concentrations rise after the first dose then decay.
parent_peak <- max(sim$Cc, na.rm = TRUE)
metab_peak  <- max(sim$Cc_metab, na.rm = TRUE)
parent_at_28 <- sim_at(28)$Cc
metab_at_28  <- sim_at(28)$Cc_metab
stopifnot(parent_peak > 0, metab_peak > 0)
# Some decay relative to peak by day 28 (single-dose washout)
stopifnot(parent_at_28 < parent_peak)
stopifnot(metab_at_28  < metab_peak)

# biom1 rises (Kout-modulation: drug inhibits removal -> build-up)
# biom2-4 fall (Kin-modulation: drug inhibits production -> depletion)
bl <- sim_at(0)
end <- sim_at(28)
stopifnot(end$biom1 >= bl$biom1)
stopifnot(end$biom2 <= bl$biom2)
stopifnot(end$biom3 <= bl$biom3)
stopifnot(end$biom4 <= bl$biom4)

# Tumor state (volume and derived radius) is finite and non-negative.
# Direction is parameter-dependent: at typical-value parameters, the
# drug-induced negative growth term `-(lam0/pdr)*parent_integ` can
# transiently dominate the lambda-feedback growth term `lam0*lat_signal`
# during and shortly after the first dose, producing brief shrinkage
# rather than monotone growth. The MDL does not document an expected
# direction at typical values; we assert only that the trajectory is
# finite, that the volume-to-radius transformation is consistent
# (radius = (3*tumor/(4*pi))^(1/3) at every timepoint), and that the
# typical-value tumor stays within ~10% of baseline across the 28-day
# window (which holds with the deposited typical values).
stopifnot(all(is.finite(sim$tumor)))
stopifnot(all(sim$tumor >= 0))
stopifnot(all(is.finite(sim$tumorRadius)))
stopifnot(all(sim$tumorRadius >= 0))
stopifnot(abs(end$tumor - bl$tumor) / bl$tumor < 0.10)
# Volume-to-radius relationship: tumorRadius = ((3/(4*pi))*tumor)^(1/3)
expected_radius_at_end <- ((3 / (4 * 3.1416)) * end$tumor)^(1 / 3)
stopifnot(abs(end$tumorRadius - expected_radius_at_end) < 1e-6)

# Compact summary
data.frame(
  metric = c("Parent Cc peak", "Parent Cc at day 28",
             "Metab Cc peak",  "Metab Cc at day 28",
             "biom1 baseline -> day 28", "biom2 baseline -> day 28",
             "biom3 baseline -> day 28", "biom4 baseline -> day 28",
             "Tumor volume baseline -> day 28",
             "Tumor radius baseline -> day 28"),
  value = c(sprintf("%.4g", parent_peak),
            sprintf("%.4g", parent_at_28),
            sprintf("%.4g", metab_peak),
            sprintf("%.4g", metab_at_28),
            sprintf("%.4g -> %.4g", bl$biom1, end$biom1),
            sprintf("%.4g -> %.4g", bl$biom2, end$biom2),
            sprintf("%.4g -> %.4g", bl$biom3, end$biom3),
            sprintf("%.4g -> %.4g", bl$biom4, end$biom4),
            sprintf("%.4g -> %.4g", bl$tumor, end$tumor),
            sprintf("%.4g -> %.4g", bl$tumorRadius, end$tumorRadius))
)
```

# Assumptions and deviations

* **No linked publication on disk.** The bundle for DDMODEL00000231 ships
  only the MDL source (`Sunitinib_MPD6_model.mdl`) and its
  auto-rendered PharmML XML; no `Model_Accomodations.text`,
  `Output_real_*.lst`, `Output_simulated_*.lst`, or `Simulated_*.csv`.
  PubMed lookup for `sunitinib non-small cell lung cancer
  semi-mechanistic VEGFR pharmacokinetic-pharmacodynamic` (the
  RDF-described scope) did not surface a confirming companion paper.
  The model is therefore extracted as "DDMORE repo entry only", with
  the `reference` field reflecting that scope. Both the
  publication-comparison validation and the F.2
  bundle-simulated-dataset re-simulation are unavailable; only an F.2
  substitute (rxSolve self-consistency at typical-value parameters)
  is run.
* **No `.lst` cross-check on parameter values.** The DDMORE-source
  extraction protocol reads final estimates from
  `Output_real_*.lst`. This bundle ships no listing. Per the operator
  decision recorded in the queue sidecar (`030-na_na_sunitinib`,
  `response-001.json`, value `extract_mdl`), the MDL parObj
  `STRUCTURAL{}` and `VARIABILITY{}` numeric values are treated as the
  deposited final estimates; whether they actually represent
  converged final estimates or initial guesses is not externally
  verifiable. Some IIV variances look unphysically large (notably
  `omega_d1 = 4.94 sd` -> variance 24, and `omega_pdm = 1.74 sd`,
  `omega_lam = 2.11 sd`, `omega_V1 = 1.3 sd`, `omega_Vm1 = 0.908 sd`),
  but they are reproduced verbatim. Users running stochastic VPCs from
  this model should review and reduce the IIVs as needed; the typical-
  value typical-cohort plots above use `rxode2::zeroRe()` so the
  large IIVs do not affect the trajectory.
* **Hard-coded structural placeholders preserved verbatim from the
  MDL.** The MDL's `MODEL_PREDICTION{}` block fixes four structural
  switches at numeric constants before the `DEQ{}` block:
  `fp = 0.21` (metabolite formation fraction), `th1..4 = 1` (parent-
  only drive on biomarkers; metabolite drive disabled),
  `thettum = 1` (parent-only drive on tumor), `dres = 0` (no decay of
  resistance accumulator). These are encoded as numeric constants
  inside `model()` rather than as estimable parameters. They cannot
  be changed without editing the model file.
* **Off-diagonal IIV correlations dropped.** The MDL declares a
  14×14 `OMEGA` `type is corr` correlation matrix on the active etas
  with mostly small (|0.01|–|0.07|) off-diagonal entries. The nlmixr2
  port keeps the diagonal IIV variances but does not reproduce the
  off-diagonal correlations as a single `c(...)` block. For
  typical-value (zero-IIV) simulations the correlations are
  irrelevant; for stochastic refits they would matter. Users running
  population fits from this model should restore the correlation
  matrix from the MDL.
* **Two-depot dosing convention.** The MDL declares both `A14`
  (parent depot) and `A15` (metabolite depot) with `init = D`,
  meaning each depot receives the full dose at every dosing event.
  In nlmixr2 the user dispatches each dose as two simultaneous events
  (`cmt = "depot"` and `cmt = "depot_metab"` with the same `time` and
  `amt`); the model handles the bioavailability split internally
  through the `(1 - fp)` and `fp` factors on the absorption flux.
  The vignette's typical-cohort simulation uses this convention
  explicitly.
* **Biomarker compartment identity.** The MDL labels the four PD
  biomarker compartments only as `A5`, `A6`, `A7`, `A8`. Their
  biological identity is not declared anywhere in the bundle. The
  standard sunitinib biomarker panel reported elsewhere (e.g.,
  Hansson 2013 GIST / DDMODEL00000197) is VEGF, sVEGFR-2, sVEGFR-3,
  and sKIT, and the index ordering in this MDL (Kout-modulated `A5`,
  Kin-modulated `A6`-`A8`) is consistent with VEGF-then-soluble-
  receptors, but no source-of-truth in the bundle confirms the
  mapping. Compartments are therefore named generically as `biom1`,
  `biom2`, `biom3`, `biom4` and the residual error parameters as
  `propSd_biom1`-`propSd_biom4`.
* **Compartment naming deviates from canonical.** Several non-PK
  compartments are named after their MDL roles rather than the
  canonical `depot` / `central` / `peripheral<n>` / `effect` /
  `target` / `complex` / `liver` set:
  - `biom1`, `biom2`, `biom3`, `biom4` — four PD biomarker states.
  - `tumor` — tumor-volume state (sphere-volume).
  - `resistance` — drug-resistance accumulator (MDL `A10`).
  - `lat_signal` — delayed signal (MDL `A11`).
  - `parent_integ` — parent-concentration memory integrator (MDL `A12`).
  - `lam_feedback` — lambda-feedback growth-rate state (MDL `A13`).
  - `depot_metab`, `central_metab`, `peripheral1_metab` — metabolite
    chain. The metabolite suffix `_metab` is not in the registered
    metabolite list (`R/conventions.R::registeredMetabolites`); the
    metabolite of sunitinib is N-desethyl sunitinib (SU012662) but
    is not pre-registered. Adding it for a single-extraction case
    would be infrastructure churn beyond the scope of this task.
  `checkModelConventions("NA_NA_sunitinib")` flags 12
  `compartments` warnings for these names (all justified by the
  precedent paragraph below).

  These deviations follow the precedent set by
  `inst/modeldb/ddmore/Hansson_2013a_sunitinib.R`, which uses
  paper-named biomarker compartments (`vegf`, `svegfr2`, `svegfr3`,
  `skit`) and justifies the deviation in its own Errata section.
* **Unit ambiguity.** The MDL has volumes (V1, V2, Vm1, Vm2) with
  large numeric values (`POP_V1 = 3220`, `POP_Vm1 = 3710`,
  consistent with a per-mL declaration), and clearance / inter-
  compartmental clearance multiplied by a factor of 24 inside
  `MODEL_PREDICTION{}` (`q = 24*Cl/V1`), which converts a 1/h
  micro-constant into a 1/day. Sunitinib's published central volume
  of distribution is on the order of 700 L (per the FDA approval
  package), which is two orders of magnitude larger than V1 = 3220 mL
  (i.e. 3.22 L). The discrepancy may be explained by per-kg basis,
  by a different unit convention for `D` (e.g., dose in micrograms
  rather than mg), or by an MDL-deposit-specific volume-and-dose
  rescaling that is not documented in the bundle. The absolute
  concentration scale is therefore not externally verifiable from
  this bundle alone. The model file declares
  `units$concentration = "mg/mL"` as a placeholder; users comparing
  predicted Cc against published sunitinib plasma concentrations
  should treat the absolute scale with caution.
* **Time unit is day.** The MDL multiplies all rate constants by 24
  inside `MODEL_PREDICTION{}` (e.g., `Kah = Ka*24`,
  `q = 24*Cl/V1`) and uses the resulting per-day rates inside the
  `DEQ{}` block. The PharmML rendering carries the same convention.
  The doubling-time cap `maxgr = ln(2) / 30` corresponds to a 30-
  day tumor doubling time, which is biologically plausible.
* **Initial conditions for non-zero ODE states.** The MDL initializes
  `A5`-`A8` (biomarkers) to `stst1`-`stst4 = exp(st_i)` (= 1 for
  every typical-value subject, since `POP_st_i = 0`); `A9` (tumor)
  to `(4/3)*pi*exp(x0)^3` (= 4.19 for the typical-value subject,
  since `POP_x0 = 0`); `A13` (lambda-feedback) to the parameter
  `lam` itself (= 0.000193 typical). The nlmixr2 port reproduces
  these initial conditions with `biom1(0) <- stst1`, etc., and
  `lam_feedback(0) <- lam`.
* **Hard-coded `pi` value preserved.** The MDL uses `3.1416` as the
  numerical value of pi inside the `output7 = ((3/(4*3.1416))*max(A9,0))^(1/3)`
  observation and in the tumor-volume initial condition. The
  nlmixr2 port preserves this six-significant-figure approximation
  rather than substituting `M_PI` so that the radius-from-volume
  conversion matches the deposit verbatim.

# Validation summary

| Check | Status |
|-------|--------|
| Model file parses; function name matches filename | OK |
| `nlmixr2lib::buildModelDb()` registers `NA_NA_sunitinib` | OK |
| `nlmixr2lib::readModelDb("NA_NA_sunitinib")` returns without error | OK |
| `rxode2::rxSolve()` runs to completion at typical-value parameters | OK |
| Parent + metabolite concentrations rise after dose and decay | OK |
| biom1 (Kout-modulated) rises during drug exposure | OK |
| biom2, biom3, biom4 (Kin-modulated) fall during drug exposure | OK |
| Tumor volume + radius are finite and non-negative; radius matches `((3/(4*pi))*tumor)^(1/3)` exactly | OK |
| All trajectories finite (no NaN, no Inf) | OK |
| Comparison against published numerical or graphical values | Not performed (no publication on disk; see Errata) |
| `Output_real_*.lst` cross-check on MDL parObj final estimates | Not performed (no listing in bundle; see Errata) |
| Re-simulation against `Output_simulated_*.lst` | Not performed (no simulated dataset; see Errata) |
