---
title: "DDMoRe: miridesap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DDMoRe: miridesap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("NA_NA_miridesap"))$reference`
* Description: `r rxode2::rxode2(readModelDb("NA_NA_miridesap"))$description`
* DDMORE Foundation Model Repository entry: [DDMODEL00000262](https://repository.ddmore.eu/model/DDMODEL00000262)
* Linked publication: **none identified** (see Errata).

This model is a population PK/PD characterisation of CPHPC (also known
as miridesap, GSK2315698, Ro 63-8695) and serum amyloid P (SAP) in
healthy volunteers and patients with systemic amyloidosis. CPHPC is a
small-molecule palindromic ligand designed to deplete plasma SAP by
binding its calcium-coordinated pentameric face; the bound CPHPC-SAP
complex is rapidly cleared from circulation by hepatocytes via FcgRn
internalization (the published clinical motivation behind the SAP-
depletion approach to systemic amyloidosis is reviewed in Pepys et al.
2002, *Nature* 417:254-259, but the corresponding popPK/PD modelling
publication for `DDMODEL00000262` could not be located on disk in this
worktree -- see Errata).

The model was extracted from the DDMORE Foundation Model Repository
bundle for `DDMODEL00000262` (scraped to
`dpastoor/ddmore_scraping/262/`). The bundle contains:

- `Executable_simulated_CPHPC_dataset.ctl` -- the NONMEM `.ctl` for the
  **base model**: 17 THETAs covering CPHPC two-compartment PK, SAP
  two-compartment turnover, CPHPC-SAP binding kinetics, an SC depot,
  and a single CRCL covariate on CPHPC clearance.
- `Output_simulated_CPHPC_dataset.lst` -- NONMEM listing from running
  the base model on `Simulated_CPHPC_dataset.csv`. The run uses
  `$EST METHOD=IMP EONLY=1` (importance-sampling **evaluation only**,
  no optimization), so the FINAL PARAMETER ESTIMATE block faithfully
  reproduces the `.ctl` initial values to within rounding -- confirming
  the `.ctl` `$THETA` block as the source of truth for the base
  model's final estimates.
- `Output_real_CPHPC.lst` -- listing for a more elaborate **FINAL
  MODEL** with 21 THETAs that adds organ-specific amyloid-load
  covariates (AMLIVER, AMSPLEEN, AMBONE, AMHEART, ...) on top of the
  base model. The corresponding `.mod` is **not** in the bundle; this
  extraction therefore stops at the base model.
- `Simulated_CPHPC_dataset.csv` -- a 38-subject simulated dataset with
  IV regimens spanning 5 mg/h x 1 h up to 70 mg/h x 1 h + 10 mg/h x 23 h
  loading-and-maintenance schedules.
- `DDMODEL00000262.rdf` -- RDF metadata: a PK/PD model
  (`pkpd_0000036`) connecting CPHPC exposure to plasma SAP. The
  description names two contributing GSK studies (CPH113776 in healthy
  volunteers and CPH114527 in amyloidosis patients) but no journal
  publication is referenced.
- `262.json`, `Command.txt` -- scraper metadata and the NONMEM run
  command.

The bundle has no `Model_Accomodations.text` file, so publication
identification rests on the RDF description and the `.ctl` `$PROB`
line (`BASE MODEL`). A PubMed search by `CPHPC` / `miridesap` / `SAP
depletion population PK` did not return a match for the specific
modelling work that produced `DDMODEL00000262`.

# Population

The DDMODEL00000262 RDF metadata names two clinical studies feeding
the model: CPH113776 (a healthy-volunteer first-in-human study of
CPHPC) and CPH114527 (an exploratory study in patients with systemic
amyloidosis). The `Output_real_CPHPC.lst` reports `TOT. NO. OF
INDIVIDUALS: 38` across the combined cohort (the `.lst` is for the
FINAL MODEL with patient-organ-load covariates, but the underlying
study population is the same). Finer-grained demographics -- age,
weight, sex distribution, region, dose escalation -- are not in the
DDMORE bundle and no linked publication was located.

The same information is available programmatically via the model's
`population` metadata (`readModelDb("NA_NA_miridesap")$population`
after the model is loaded).

# Source trace

The per-parameter origin is recorded inline in
`inst/modeldb/ddmore/NA_NA_miridesap.R`. The table below collects them
in one place for review. **All structural-parameter values come from
the bundle's `Executable_simulated_CPHPC_dataset.ctl`; the
`Output_simulated_CPHPC_dataset.lst` reproduces those values as the
"final estimates" of an `EONLY=1` evaluation run.** No linked
publication was identified for cross-validation.

| nlmixr2 parameter | Source THETA | Source value | Linear value | Source location |
|-------------------|-------------|-------------|-------------|-----------------|
| `lcl`             | THETA(2)    | 1.92        | CL = 6.82 L/h | `.ctl` `$THETA` line 201 (`; 2. CL`) |
| `lvc`             | THETA(3)    | 2.76        | Vc = 15.8 L | `.ctl` `$THETA` line 202 (`; 3. V`) |
| `lq`              | THETA(4)    | 0.595       | Q = 1.81 L/h | `.ctl` `$THETA` line 203 (`; 4. Q`) |
| `lvp`             | THETA(5)    | 2.84        | Vp = 17.1 L | `.ctl` `$THETA` line 204 (`; 5. V2`) |
| `e_crcl_cl`       | THETA(15)   | 0.0152      | -            | `.ctl` `$THETA` line 214 (`; 15. CRCLCOV`) |
| `lka`  (FIXED)    | THETA(14)   | 0.4055      | ka = 1.50 1/h | `.ctl` `$THETA` line 213 (`; 14. KASC`) |
| `lfdepot` (FIXED) | THETA(13)   | 1 (linear)   | F = 1        | `.ctl` `$THETA` line 212 (`; 13. FSC`) |
| `lkout`           | THETA(1)    | -3.07       | kout = 0.0464 1/h | `.ctl` `$THETA` line 200 (`; 1. KOUT`) |
| `lsap0`           | THETA(6)    | 3.37        | sap0 = 29.1 mg/L | `.ctl` `$THETA` line 205 (`; 6. SAP10`) |
| `lvp_sap`         | THETA(16)   | 3.47        | Vp_sap = 32.1 L | `.ctl` `$THETA` line 215 (`; 16. V4`) |
| `lq_sap`          | THETA(17)   | 1.49        | Q_sap = 4.44 L/h | `.ctl` `$THETA` line 216 (`; 17. Q4`) |
| `lkon`            | THETA(7)    | 14.6        | kon = 2.19e6 L/(mol h) | `.ctl` `$THETA` line 206 (`; 7. KON`) |
| `lkint`           | THETA(8)    | 5.71 (linear) | kint = 5.71 1/h | `.ctl` `$THETA` line 207 (`; 8. KINT`) |
| `propSd`          | THETA(9)    | 0.286       | -            | `.ctl` `$THETA` line 208 (`; 9. PROPPK`) |
| `propSd_sap`      | THETA(11)   | 0.268       | -            | `.ctl` `$THETA` line 210 (`; 11. PROPSAP`) |
| `etalkout`        | OMEGA(1)    | 0.125       | -            | `.ctl` `$OMEGA` line 217 (`; 1. OMKOUT`) |
| `etalcl`          | OMEGA(2)    | 0.0474      | -            | `.ctl` `$OMEGA` line 218 (`; 2. OMCL`) |
| `etalvc`          | OMEGA(3)    | 0.0836      | -            | `.ctl` `$OMEGA` line 219 (`; 3. OMV`) |
| `etalq`  (FIXED)  | OMEGA(4)    | 0.0225      | -            | `.ctl` `$OMEGA` line 220 (`; 4. OMQ`) |
| `etalvp` (FIXED)  | OMEGA(5)    | 0.0225      | -            | `.ctl` `$OMEGA` line 221 (`; 5. OMV2`) |
| `etalsap0`        | OMEGA(6)    | 0.0625      | -            | `.ctl` `$OMEGA` line 222 (`; 6. OMSAP10`) |
| `etalkon`  (FIXED)| OMEGA(7)    | 0.0225      | -            | `.ctl` `$OMEGA` line 223 (`; 7. OMKON`) |
| `etalkint` (FIXED)| OMEGA(8)    | 0.0225      | -            | `.ctl` `$OMEGA` line 224 (`; 8. OMKINT`) |
| `etalq_sap`       | OMEGA(9)    | 1.11        | -            | `.ctl` `$OMEGA` line 225 (`; 9. OMQ4`) |
| `etalvp_sap`      | OMEGA(10)   | 2.05        | -            | `.ctl` `$OMEGA` line 226 (`; 10. OMV4`) |
| `etalka` (FIXED)  | OMEGA(11)   | 0.0225      | -            | `.ctl` `$OMEGA` line 227 (`; 11. OMKSC`) |
| `MOLCPH`          | constant    | 340.37 kDa  | mg/mol       | `.ctl` `$PK` line 40 |
| `MOLSAP`          | constant    | 5 x 25 kDa  | mg/mol       | `.ctl` `$PK` line 39 |
| `d/dt(depot)`     | -           | -           | -            | `.ctl` `$DES` line 163 (`-KSC*A5`) + dosing into A5 |
| `d/dt(central)`   | -           | -           | -            | `.ctl` `$DES` line 158 (CPHPC mass balance) |
| `d/dt(peripheral1)` | -         | -           | -            | `.ctl` `$DES` line 159 (CPHPC peripheral) |
| `d/dt(total_target)` | -        | -           | -            | `.ctl` `$DES` line 161 (SAP central total) |
| `d/dt(target_peripheral)` | -   | -           | -            | `.ctl` `$DES` line 162 (SAP peripheral free) |
| `d/dt(complex)`   | -           | -           | -            | `.ctl` `$DES` line 164 (CPHPC-SAP complex) |
| Observation `Cc`  | -           | scaling     | ng/mL        | `.ctl` `$PK` line 135 (`S1 = V/1000`) |
| Observation `sap` | -           | scaling     | ng/mL-equiv  | `.ctl` `$PK` line 136 (`S3 = 1/(1000*MOLSAP)`) |

Source-trace note on the `target_peripheral` compartment name: see
[Errata](#errata).

# Virtual cohort

The bundle's `Simulated_CPHPC_dataset.csv` carries 38 simulated
subjects with IV-only regimens. To exercise the full structural model
- including the SC depot which the bundle's simulated dataset does not
exercise - we build a small representative virtual cohort below
(typical-value subjects covering normal renal function and a CRCL = 50
mL/min subject to demonstrate the CRCL covariate effect).

```{r cohort}
set.seed(20260506)

# Three subjects: two with normal CRCL (one IV-only, one IV + SC), and
# one with reduced renal function on IV-only.
subjects <- tibble::tribble(
  ~id, ~CRCL, ~regimen,           ~scenario,
  1L,  100,   "iv_loading_main",  "IV 70 mg/h x1h + 10 mg/h x23h, CRCL 100",
  2L,   50,   "iv_loading_main",  "IV 70 mg/h x1h + 10 mg/h x23h, CRCL 50",
  3L,  100,   "sc_single",        "SC depot 200 mg, CRCL 100"
)

obs_times <- sort(unique(c(seq(0, 36, by = 0.25))))

make_cohort <- function(subjects, id_offset = 0L) {
  out <- list()
  for (i in seq_len(nrow(subjects))) {
    s <- subjects[i, ]
    if (s$regimen == "iv_loading_main") {
      ev <- rxode2::et(amt = 70, rate = 70, cmt = "central", time = 0) |>
        rxode2::et(amt = 230, rate = 10, cmt = "central", time = 1) |>
        rxode2::et(obs_times, cmt = "Cc") |>
        rxode2::et(obs_times, cmt = "sap")
    } else if (s$regimen == "sc_single") {
      ev <- rxode2::et(amt = 200, cmt = "depot", time = 0) |>
        rxode2::et(obs_times, cmt = "Cc") |>
        rxode2::et(obs_times, cmt = "sap")
    } else {
      stop("unknown regimen: ", s$regimen)
    }
    df <- as.data.frame(ev)
    df$id       <- id_offset + s$id
    df$CRCL     <- s$CRCL
    df$scenario <- s$scenario
    out[[i]]    <- df
  }
  dplyr::bind_rows(out)
}

events <- make_cohort(subjects)

# Cheap regression guard against accidental id collisions.
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod <- rxode2::zeroRe(readModelDb("NA_NA_miridesap"))

sim <- rxode2::rxSolve(
  mod, events = events,
  keep = c("CRCL", "scenario"),
  returnType = "data.frame", addDosing = FALSE
) |>
  dplyr::filter(!is.na(time))
```

For demonstration of between-subject variability in the SAP-recovery
phase (where the IIV is non-trivial), we also run a small stochastic
cohort - 50 subjects all on the IV loading + maintenance regimen at
CRCL = 100:

```{r vpc-cohort}
mod_iiv <- readModelDb("NA_NA_miridesap")  # keeps IIV active
n_vpc <- 50L
vpc_events <- make_cohort(
  tibble::tibble(
    id       = seq_len(n_vpc),
    CRCL     = rep(100, n_vpc),
    regimen  = rep("iv_loading_main", n_vpc),
    scenario = rep("VPC: IV 70 mg/h + 10 mg/h, CRCL 100", n_vpc)
  )
)
sim_vpc <- rxode2::rxSolve(
  mod_iiv, events = vpc_events, nSub = 1L,
  keep = c("CRCL", "scenario"),
  returnType = "data.frame", addDosing = FALSE
) |>
  dplyr::filter(!is.na(time))
```

# Replicate trajectories

The bundle's `Output_simulated_CPHPC_dataset.lst` does not ship its
`$TABLE` predictions in a format we can plot directly here, but the
key qualitative behaviours of the base model are: (a) CPHPC plasma
concentration rises during the loading dose to a few thousand ng/mL,
sustains during the maintenance infusion, then declines biexponentially
once dosing stops; (b) plasma SAP drops by 1-2 orders of magnitude
during dosing - the published clinical mechanism of SAP depletion -
and recovers towards baseline at the endogenous turnover half-life
(`ln(2) / kout` ~ `r round(log(2) / 0.0464, 1)` h); (c) renal
impairment reduces CPHPC clearance, raising plasma exposure.

```{r figure-cphpc}
sim |>
  ggplot(aes(time, Cc, colour = scenario, linetype = scenario)) +
  geom_line(linewidth = 0.6) +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "CPHPC plasma concentration (ng/mL)",
    colour = NULL, linetype = NULL,
    title = "CPHPC plasma trajectory by regimen / CRCL",
    caption = "Typical-value (zero-IIV) simulation of NA_NA_miridesap."
  ) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(ncol = 1))
```

```{r figure-sap}
sim |>
  ggplot(aes(time, sap / 1000, colour = scenario, linetype = scenario)) +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = exp(3.37), linetype = "dotted", colour = "grey40") +
  scale_y_continuous(trans = "log10") +
  labs(
    x = "Time (h)",
    y = "Plasma SAP (mg/L; horizontal dotted line = baseline 29.1 mg/L)",
    colour = NULL, linetype = NULL,
    title = "Plasma SAP trajectory: depletion during CPHPC dosing and recovery",
    caption = "Typical-value (zero-IIV) simulation of NA_NA_miridesap. SAP is converted ng/mL -> mg/L by /1000."
  ) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(ncol = 1))
```

```{r figure-vpc-sap}
sim_vpc |>
  group_by(time) |>
  summarise(
    q05 = quantile(sap, 0.05, na.rm = TRUE),
    q50 = quantile(sap, 0.50, na.rm = TRUE),
    q95 = quantile(sap, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, q50 / 1000)) +
  geom_ribbon(aes(ymin = q05 / 1000, ymax = q95 / 1000), alpha = 0.2) +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = exp(3.37), linetype = "dotted", colour = "grey40") +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Plasma SAP (mg/L)",
    title = "Stochastic VPC: plasma SAP under IV loading + maintenance",
    caption = sprintf(
      "n = %d subjects, IV 70 mg/h x1h + 10 mg/h x23h, CRCL = 100 mL/min. Median + 90%% prediction interval.",
      n_vpc
    )
  ) +
  theme_bw()
```

# Self-consistency check (F.2)

There is no linked publication to compare against; per the
`extract-literature-model` skill's `references/ddmore-source.md`
decision tree, validation reduces to (a) the model parses, (b)
`rxode2::rxSolve()` runs to completion on a representative regimen,
(c) qualitative behaviours match the source `.ctl`'s known dynamics,
(d) typical-value values stay in physiologically plausible ranges.

```{r f2-self-consistency}
sim_unique <- sim |>
  dplyr::distinct(scenario, time, .keep_all = TRUE)

sanity <- sim_unique |>
  dplyr::group_by(scenario) |>
  dplyr::summarise(
    cphpc_baseline_ng_mL  = Cc[which(time == min(time))[1]],
    cphpc_peak_ng_mL      = max(Cc, na.rm = TRUE),
    cphpc_final_ng_mL     = Cc[which.max(time)],
    sap_baseline_mg_L     = sap[which(time == min(time))[1]] / 1000,
    sap_min_mg_L          = min(sap, na.rm = TRUE) / 1000,
    sap_recovery_t30_mg_L = {
      idx <- which.min(abs(time - 30))
      sap[idx] / 1000
    },
    .groups = "drop"
  )
knitr::kable(sanity, digits = 2,
             caption = paste("Per-scenario sanity summary. CPHPC baseline = 0",
                             "(no drug at t=0); SAP baseline ~ 29 mg/L (= exp(3.37));",
                             "CPHPC peak driven by the loading dose; SAP minimum",
                             "reached during sustained CPHPC exposure."))
```

The numbers above are the F.2 acceptance criteria: CPHPC baseline = 0,
SAP baseline matches `exp(lsap0) = 29.1 mg/L` (small numeric drift OK
because the proportional scaling carries through `MOLSAP * 1000`),
SAP minimum is well below baseline (large depletion during sustained
exposure - the clinical mechanism of CPHPC), and reduced CRCL increases
CPHPC exposure (the CRCL = 50 scenario shows a higher final
concentration than CRCL = 100 at the same dose, since `crcl_effect =
1 + 0.0152 * (50 - 80) = 0.544` reduces CL to 54% of typical).

# Errata, assumptions, and deviations

- **No linked publication.** A PubMed / Google Scholar search by
  `CPHPC` / `miridesap` / `SAP depletion population PK / PK-PD` did
  not return a journal publication corresponding to
  `DDMODEL00000262`. The `Pepys et al. 2002 Nature` paper (cited in
  the model description) describes the drug discovery / clinical
  mechanism but not the pop PK/PD modelling work. The DDMORE bundle
  has no `Model_Accomodations.text` file. The `reference` field
  therefore records "DDMORE Foundation Model Repository: DDMODEL00000262.
  No linked publication."
- **Base model rather than FINAL MODEL.** The bundle ships the `.ctl`
  for the BASE MODEL (17 THETAs; CRCL covariate only) but the `.lst`
  for the FINAL MODEL (21 THETAs; CRCL plus organ-specific
  amyloid-load covariates AMLIVER, AMSPLEEN, AMBONE, AMHEART, ...).
  We extract the base model because the FINAL MODEL's `.mod` is not in
  the bundle - without it, the structural code for the four extra
  THETAs (`TH18` ... `TH21`) is unrecoverable. A future extraction
  could add a `_final` variant if the operator obtains the FINAL
  MODEL's source `.mod`.
- **Final estimates source.** The `.ctl`'s `$THETA` initial values are
  the source of truth: `Output_simulated_CPHPC_dataset.lst` reproduces
  them exactly because the simulated run is `$EST METHOD=IMP EONLY=1`
  (importance-sampling **evaluation only**, no optimization). The
  Output_real_CPHPC.lst's FINAL PARAMETER ESTIMATE block reports
  slightly different values for the structural THETAs (e.g.
  `lvc = 2.78` vs `.ctl` 2.76, `lq = 0.542` vs 0.595) because that
  listing is for the FINAL MODEL with covariates re-fit on the real
  data; we follow the bundle's executable `.ctl` rather than the
  FINAL MODEL's listing because only the `.ctl` ships re-runnable code.
- **Bundle "FSC = 1 FIXED".** The `.ctl`'s SC bioavailability THETA(13)
  is fixed at 1, so subcutaneous and IV doses produce identical AUC
  for a given mg dose. This is unusual for a real SC formulation and
  may reflect a placeholder choice in the modelling work - the bundle
  has no documentation explaining it. We propagate the FIX as-is.
- **`target_peripheral` compartment naming deviation.** The model
  carries an SAP peripheral compartment whose canonical-naming
  analogue in nlmixr2lib (`target_peripheral`) is not in the
  `R/conventions.R` registered compartment list. `checkModelConventions()`
  flags it as a warning. The deviation is justified: the source model
  has a target (SAP) with its own central + peripheral distribution,
  and the registered names (`target` for free, `complex` for bound,
  `total_target` for total under QSS / MM TMDD) do not cover the
  "free target peripheral compartment" concept. Adding the new name
  to the conventions register is out of scope for this PR; if more
  models with the same shape are added in the future the convention
  should be extended rather than each model carrying the deviation
  independently.
- **`KOFF` is computed in the `.ctl` but not used in the `$DES`.** The
  bundle defines `KD = 10 nM` and `KOFF = KON * KD` (lines 101-104) but
  `DADT(6)` (the bound-complex ODE) has only `KON*C3F*C1F - KINT*C3C1`
  with no `+ KOFF * C3C1` term - i.e. the binding is treated as
  effectively irreversible because internalization is much faster than
  dissociation (`KINT = 5.71 / h` versus `KOFF = 0.0219 / h`). We omit
  KD / KOFF from the nlmixr2 model for the same reason; the simulation
  trajectories below should be identical to a model with KOFF
  explicitly carried because of the order-of-magnitude separation.
- **Mixed-units compartments.** Following the source `.ctl`, the
  CPHPC compartments (`depot`, `central`, `peripheral1`) carry mass in
  mg, while the SAP / complex compartments (`total_target`,
  `target_peripheral`, `complex`) carry concentration in mol/L. Unit
  conversions between the two systems use the molecular weights
  `MOLCPH = 340.37 kDa` (CPHPC) and `MOLSAP = 5 * 25 kDa` (SAP
  pentamer). Reviewers verifying dimensional consistency should pay
  particular attention to the mixed-units terms in `d/dt(central)`
  (the `kint * vc * MOLCPH * complex` term converts mol/L to mg/h)
  and `d/dt(target)` (the `k43 * target_peripheral * vp_sap / vc`
  term carries volume ratios because peripheral free SAP is moved as
  concentration into the central total compartment).
- **Dosing units / scaling carry-through.** The `.ctl`'s `S1 = V / 1000`
  and `S3 = 1 / (1000 * MOLSAP)` convert mass and concentration to
  ng/mL for both observations. We propagate this unchanged:
  `Cc = central / vc * 1000` (mg/L * 1000 = ng/mL) and
  `sap = total_target * MOLSAP * 1000` (mol/L * mg/mol * 1000 = ng/mL =
  mg/L * 1000). The SAP observation column is therefore on a
  ng/mL-equivalent scale; figures and tables in this vignette divide
  by 1000 to express SAP in the conventional clinical mg/L unit.
- **DDMORE "FINAL MODEL" listing reports `R MATRIX ALGORITHMICALLY
  SINGULAR`-adjacent diagnostics.** The Output_real_CPHPC.lst notes
  "Number of Negative Eigenvalues in Matrix=20 ... Forcing positive
  definiteness" (line 327-330), suggesting the FINAL MODEL's covariance
  step did not converge cleanly. We extract the BASE MODEL only, and
  the simulated-data `.lst` (which is for the BASE MODEL) reports a
  clean importance-sampling evaluation, so this advisory affects the
  un-extracted FINAL MODEL only.
