---
title: "Afatinib (Nakao 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Afatinib (Nakao 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Nakao_2019_afatinib"))$reference`
* Description: `r rxode2::rxode(readModelDb("Nakao_2019_afatinib"))$description`
* Article: <https://doi.org/10.1038/s41598-019-54804-9> (Sci Rep 2019, open access)
* Supplement: Supplementary Information (Tables S1-S2, Figures S1-S2), available from the article page.

# Population

Nakao et al. (2019) fit the model to 354 afatinib plasma concentrations from
34 Japanese adults with EGFR mutation-positive advanced non-small cell lung
cancer (all adenocarcinoma) who started afatinib 40 mg once daily at two
Osaka centres between August 2014 and May 2016 (Table 1). Twenty-three were
female (67.6%). Age was 45-86 years (mean 66.8), body weight 35.5-79.1 kg
(mean 53.8), BMI 15.2-28.1 kg/m^2 (mean 21.9), AST 13-65 IU/L (mean 25.6),
and Cockcroft-Gault creatinine clearance 42.3-131.8 mL/min (mean 80.8).
Twenty-four had received prior chemotherapy and 21 a prior EGFR TKI. Plasma
was sampled at 0.5-1, 2-3, 4-6, 8-12 and 24 h after the first dose (day 1)
and pre-dose through 24 h on day 8, and assayed by HPLC. The model was fit in
Phoenix NLME 7.0 with FOCE-ELS.

The same information is available programmatically via
`readModelDb("Nakao_2019_afatinib")()$population`.

# Source trace

Every `ini()` value carries an in-file comment in
`inst/modeldb/specificDrugs/Nakao_2019_afatinib.R`; the table collects them.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Structure: 1-compartment, first-order absorption and elimination | -- | Results paragraph after Table 2; Methods 'Development of population pharmacokinetics model' |
| `CL/F = thetaCL * (1 + (Ccr - 79.9) * thetaCcr) * (1 + (AST - 25.3) * thetaAST)` | -- | Table 2 |
| `V/F = thetaV * (1 + (BMI - 21.8) * thetaBMI) * (1 + (Age - 66.7) * thetaAge)` | -- | Table 2 |
| `lka` | log(0.60) 1/h | Table 2 |
| `lcl` | log(20.0) L/h | Table 2 |
| `lvc` | log(795.8) L | Table 2 |
| `e_crcl_cl` | 0.0013 min/mL | Table 2 |
| `e_ast_cl` | -0.016 L/IU | Table 2 |
| `e_bmi_vc` | 0.019 m^2/kg | Table 2 |
| `e_age_vc` | -0.004 1/year | Table 2 |
| `etalka` | 0.62221 = log(1 + 0.929^2) | Table 2, omega ka 92.9% |
| `etalcl` | 0.46169 = log(1 + 0.766^2) | Table 2, omega CL/F 76.6% |
| `etalvc` | 0.24591 = log(1 + 0.528^2) | Table 2, omega V/F 52.8% |
| `propSd` | 0.317 | Table 2, sigma 31.7% |
| Exponential IIV | -- | Methods 'Inter-individual variability was modelled exponentially' |

# Virtual cohort

Table 1 prints each continuous covariate as "mean +/- SD", but the dispersions
(e.g. age 66.8 +/- 1.5 over a 45-86 year range) are only consistent with the
standard error of the mean. The cohort below therefore uses
`SD = SEM * sqrt(34)` and truncates each draw to the observed Table 1 range.
AST is right-skewed (13-65 IU/L around a mean of 25.6) and is drawn
log-normally.

```{r cohort}
rxode2::rxSetSeed(20191206)
n_sub <- 200

draw_trunc <- function(n, mean, sd, lo, hi) {
  pmin(pmax(rnorm(n, mean, sd), lo), hi)
}

cohort <- data.frame(
  id = seq_len(n_sub),
  AGE = draw_trunc(n_sub, 66.8, 1.5 * sqrt(34), 45, 86),
  BMI = draw_trunc(n_sub, 21.9, 0.5 * sqrt(34), 15.2, 28.1),
  CRCL = draw_trunc(n_sub, 80.8, 3.8 * sqrt(34), 42.3, 131.8),
  AST = pmin(pmax(exp(rnorm(n_sub, log(23.5), 0.40)), 13), 65)
)
summary(cohort[, -1])
```

# Simulation

Afatinib 40 mg once daily for 8 days, with dense sampling on day 1 (0-24 h)
and day 8 (168-192 h), matching the paper's sampling days.

```{r events}
mod <- readModelDb("Nakao_2019_afatinib")

obs_times <- c(0, 0.5, 1, 2, 3, 4, 5, 6, 8, 10, 12, 16, 20, 24)
obs_grid <- sort(unique(c(obs_times, 168 + obs_times)))

dose_rows <- data.frame(
  time = seq(0, 168, by = 24), amt = 40, evid = 1, cmt = "depot"
)
obs_rows <- data.frame(time = obs_grid, amt = 0, evid = 0, cmt = "central")
ev_one <- dplyr::bind_rows(dose_rows, obs_rows) |>
  dplyr::arrange(time, dplyr::desc(evid))

events <- tidyr::crossing(id = cohort$id, ev_one) |>
  dplyr::left_join(cohort, by = "id") |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

sim <- rxode2::rxSolve(mod, events, keep = c("AGE", "BMI", "CRCL", "AST")) |>
  as.data.frame()
```

## Typical-value check against the closed form

With the random effects zeroed, the model for the typical patient
(age 66.7 years, BMI 21.8 kg/m^2, AST 25.3 IU/L, CrCl 79.9 mL/min) must
reproduce the analytic one-compartment oral solution with CL/F = 20.0 L/h and
V/F = 795.8 L. Both sides use identical parameters, so the difference is
purely numerical and a tight bound is correct.

```{r typical}
ref_pt <- data.frame(AGE = 66.7, BMI = 21.8, CRCL = 79.9, AST = 25.3)
ev_typ <- dplyr::bind_rows(dose_rows, obs_rows) |>
  dplyr::arrange(time, dplyr::desc(evid)) |>
  dplyr::mutate(id = 1L) |>
  cbind(ref_pt)
typ <- rxode2::rxSolve(rxode2::zeroRe(mod), ev_typ) |> as.data.frame()

ka <- 0.60
kel <- 20.0 / 795.8
closed <- function(t) {
  tt <- outer(t, seq(0, 168, by = 24), "-")
  tt[tt < 0] <- NA
  contrib <- 40 * ka / (795.8 * (ka - kel)) * (exp(-kel * tt) - exp(-ka * tt))
  1000 * rowSums(contrib, na.rm = TRUE)
}
chk <- data.frame(time = typ$time, sim = typ$Cc, closed = closed(typ$time))
chk <- chk[chk$time != 0, ]
max_rel_err <- max(abs(chk$sim - chk$closed) / chk$closed)
max_rel_err
stopifnot(max_rel_err < 1e-4)

# Typical half-life (paper: 37 h at steady state from the label; NCA day 1
# geometric mean 27.6 h, Table S1)
log(2) / kel
```

# Replicate published figures

## Figure 3 -- effect of AST on steady-state exposure

Replicates Figure 3 of Nakao 2019: median and 95% prediction interval of
afatinib concentrations after 40 mg once daily for 8 days in the typical
reference patient (age 66.7 years, BMI 21.8 kg/m^2, CrCl 79.9 mL/min) at AST
13.0, 25.3 and 65.0 IU/L.

```{r fig3, fig.width = 7, fig.height = 4}
fine_obs <- data.frame(time = seq(0, 192, by = 1), amt = 0, evid = 0, cmt = "central")
ev_fine <- dplyr::bind_rows(dose_rows, fine_obs) |>
  dplyr::arrange(time, dplyr::desc(evid))

scenario_events <- function(covs, label, n = 200) {
  tidyr::crossing(id = seq_len(n), ev_fine) |>
    dplyr::mutate(
      AGE = covs$AGE, BMI = covs$BMI, CRCL = covs$CRCL, AST = covs$AST,
      scenario = label
    ) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    as.data.frame()
}

sim_scenario <- function(covs, label) {
  ev <- scenario_events(covs, label)
  rxode2::rxSolve(mod, ev, keep = "scenario") |>
    as.data.frame() |>
    dplyr::group_by(scenario, time) |>
    dplyr::summarise(
      median = stats::median(Cc),
      lo = stats::quantile(Cc, 0.025),
      hi = stats::quantile(Cc, 0.975),
      .groups = "drop"
    )
}

base_cov <- list(AGE = 66.7, BMI = 21.8, CRCL = 79.9, AST = 25.3)
fig3 <- dplyr::bind_rows(
  sim_scenario(modifyList(base_cov, list(AST = 13.0)), "AST = 13.0 IU/L"),
  sim_scenario(base_cov, "AST = 25.3 IU/L"),
  sim_scenario(modifyList(base_cov, list(AST = 65.0)), "AST = 65.0 IU/L")
)

ggplot(fig3, aes(time, median)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "steelblue", alpha = 0.25) +
  geom_line(colour = "steelblue4") +
  facet_wrap(~scenario) +
  labs(x = "Time (h)", y = "Afatinib (ng/mL)",
       caption = "Replicates Figure 3 of Nakao 2019")
```

The typical-value factor on CL/F at AST 65 IU/L is
`1 + (65 - 25.3) * -0.016 = 0.365`, i.e. a 2.7-fold increase in steady-state
exposure, which is the "significant effect" of hepatic impairment the paper
reports.

## Figure 4 -- CrCl, age and BMI at the extremes of the cohort

Replicates Figure 4 of Nakao 2019 (each covariate moved to the observed
Table 1 minimum and maximum, others held at reference). The paper concludes
these covariates do not notably change exposure; the simulated medians agree.

```{r fig4, fig.width = 7, fig.height = 6}
fig4 <- dplyr::bind_rows(
  sim_scenario(base_cov, "Reference"),
  sim_scenario(modifyList(base_cov, list(CRCL = 42.3)), "CrCl = 42.3 mL/min"),
  sim_scenario(modifyList(base_cov, list(CRCL = 131.8)), "CrCl = 131.8 mL/min"),
  sim_scenario(modifyList(base_cov, list(AGE = 45)), "Age = 45 years"),
  sim_scenario(modifyList(base_cov, list(AGE = 86)), "Age = 86 years"),
  sim_scenario(modifyList(base_cov, list(BMI = 15.2)), "BMI = 15.2 kg/m^2"),
  sim_scenario(modifyList(base_cov, list(BMI = 28.1)), "BMI = 28.1 kg/m^2")
)

ggplot(fig4, aes(time, median)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "darkorange", alpha = 0.25) +
  geom_line(colour = "darkorange4") +
  facet_wrap(~scenario, ncol = 3) +
  labs(x = "Time (h)", y = "Afatinib (ng/mL)",
       caption = "Replicates Figure 4 of Nakao 2019")
```

## Figure 5 -- patient #34 with hepatic impairment

Figure 5 simulates patient #34 (age 63 years, BMI 22.3 kg/m^2, AST 65.0
IU/L, CrCl 101.8 mL/min) with the post hoc estimates ka 0.58 1/h,
CL/F 3.1 L/h, V/F 465.7 L. The typical values implied by the covariate model
for this patient are printed for context; the individual profile is then
solved with the post hoc values plugged in directly (covariates at the
reference values so every covariate factor equals 1).

```{r fig5, fig.width = 6, fig.height = 4}
pt34 <- list(AGE = 63, BMI = 22.3, CRCL = 101.8, AST = 65.0)
tv_cl34 <- 20.0 * (1 + (pt34$CRCL - 79.9) * 0.0013) * (1 + (pt34$AST - 25.3) * -0.016)
tv_v34 <- 795.8 * (1 + (pt34$BMI - 21.8) * 0.019) * (1 + (pt34$AGE - 66.7) * -0.004)
c(typical_CL = tv_cl34, typical_V = tv_v34)

mod34 <- rxode2::zeroRe(mod) |>
  rxode2::ini(lka = log(0.58), lcl = log(3.1), lvc = log(465.7))
ev34 <- ev_fine |>
  dplyr::mutate(id = 1L) |>
  cbind(ref_pt)
sim34 <- rxode2::rxSolve(mod34, ev34) |>
  as.data.frame()
stopifnot(
  abs(sim34$cl[1] - 3.1) < 1e-6,
  abs(sim34$vc[1] - 465.7) < 1e-6
)

ggplot(sim34, aes(time, Cc)) +
  geom_line() +
  labs(x = "Time (h)", y = "Afatinib (ng/mL)",
       caption = "Replicates Figure 5 of Nakao 2019 (simulated curve only)")
```

With CL/F of 3.1 L/h the accumulation ratio is large and the concentration
is still rising at day 8, which is the elevation the paper highlights.

# PKNCA validation

NCA over the day 1 (0-24 h) and day 8 (168-192 h) dosing intervals of the
stochastic cohort.

```{r nca}
conc <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(day = ifelse(time < 168, "Day 1", "Day 8"),
                treatment = "40 mg QD") |>
  dplyr::select(id, time, Cc, day, treatment)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::mutate(treatment = "40 mg QD") |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)
intervals <- data.frame(
  start = c(0, 168), end = c(24, 192),
  cmax = TRUE, tmax = TRUE, auclast = TRUE, half.life = TRUE
)
nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_res <- as.data.frame(nca$result) |>
  dplyr::mutate(day = ifelse(start < 168, "Day 1", "Day 8")) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "half.life"))
```

## Comparison against published NCA

Supplementary Table S1 reports geometric means (median for Tmax) of the
observed NCA. `ncaComparisonTable()` summarises the simulated cohort by its
median, which for these log-normally distributed quantities is the geometric
mean.

```{r nca-compare}
reference <- data.frame(
  day = c("Day 1", "Day 8"),
  cmax = c(52.6, 106.4),
  tmax = c(4.1, 3.0),
  auclast = c(816.8, 1983.1),
  half.life = c(27.6, 38.3)
)
cmp <- nlmixr2lib::ncaComparisonTable(
  nca_res, reference, by = "day",
  units = c(cmax = "ng/mL", tmax = "h", auclast = "ng*h/mL", half.life = "h")
)
knitr::kable(cmp)
attr(cmp, "footnote")
```

```{r nca-assert}
sim_auc <- nca_res |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::group_by(day) |>
  dplyr::summarise(med = stats::median(PPORRES), .groups = "drop")
ratio <- sim_auc$med / reference$auclast
ratio
# Structural gate on the centre of the distribution: a mis-transcribed CL/F,
# V/F, dose or unit moves the median AUC by far more than 30%.
stopifnot(all(abs(ratio - 1) < 0.30))
```

Day 1 and day 8 Cmax and AUC0-24 agree with the observed geometric means.
The NCA half-life from a single 24-h window is a poor estimate of a ~27 h
terminal half-life for either the observed or simulated data, and the
observed day 8 value (38.3 h) also reflects the smaller day 8 CL/F seen in
Table S1 (6.6 L/h vs 20.9 L/h on day 1), which a time-invariant model does
not reproduce; discrepancies in the half-life rows are expected and are not
tuned.

# Assumptions and deviations

* **IIV scale.** Table 2 prints omega as a percentage without stating the
  conversion. It is interpreted as a log-normal CV%, so
  `omega^2 = log(1 + CV^2)`. Supporting evidence: treating the bootstrap
  2.5-97.5th percentiles as a symmetric interval on the variance scale, the
  implied RSE of the variance is 30.5% for ka (printed CV% 30.3) and 19.6% for
  CL/F (printed 18.2) under this reading, versus 45% and 24% if the percentage
  were `100 * sqrt(omega^2)`. The V/F row does not discriminate cleanly.
  If the percentages are instead `100 * sqrt(omega^2)`, the variances would be
  0.863, 0.587 and 0.279.
* **Residual error.** `sigma (%) = 31.7` is encoded as a proportional error
  (SD 0.317). The paper states that additive, proportional, combined and power
  models were tested but does not name the final one; a percentage-valued
  sigma implies proportional.
* **Covariate centring** uses the values printed in the Table 2 equations
  (79.9 mL/min, 25.3 IU/L, 21.8 kg/m^2, 66.7 years), which differ slightly from
  the Table 1 means (80.8, 25.6, 21.9, 66.8).
* **Linear covariate factors** are encoded as published. The AST factor on
  CL/F reaches zero at AST ~ 87.8 IU/L (above the observed maximum of 65 IU/L);
  the model should not be used for AST beyond the observed range.
* **Table 1 dispersions** are labelled SD but are numerically SEMs; the
  virtual cohort uses `SEM * sqrt(34)` truncated to the observed ranges, and
  draws AST log-normally. The simulated cohort is a plausible approximation of
  the study population, not a reconstruction of it.
* **Bioavailability** is not identifiable from oral-only data; CL and V are
  apparent (CL/F, V/F) and no `fdepot` term is included.
* No errata for the article were found (checked 2026-09-24).
