---
title: "Valproic acid seizure-control exposure-response (Nakashima 2015)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Valproic acid seizure-control exposure-response (Nakashima 2015)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Nakashima 2015 asks a therapeutic-drug-monitoring question: given that valproic
acid (VPA) plasma concentrations correlate well with dose but poorly with
seizure control, what is the *optimal* trough concentration for an individual
patient? The answer is a logistic exposure-response model for the probability of
an over-50% reduction in seizure frequency, driven by the model-predicted
steady-state trough VPA concentration, from which an optimal trough can be
back-solved per patient at a ROC-derived cut-off.

The paper fits **three** such models, and this package carries all three as
separate model files, per the replicate-the-author's-structure policy:

| Model file | Source | Population |
|---|---|---|
| `Nakashima_2015_valproic_acid` | Eq 6 + Table 3 | all ages, n = 77 (the paper's final model) |
| `Nakashima_2015_valproic_acid_le18y` | Eq 7 | aged 18 y or younger, n = 56 |
| `Nakashima_2015_valproic_acid_ge19y` | Eq 8 | aged 19 y or older, n = 21 |

* Article: <https://doi.org/10.1371/journal.pone.0141266>
* Expression of Concern: <https://doi.org/10.1371/journal.pone.0279487>
* Upstream population PK source (Ogusu 2014): <https://doi.org/10.1371/journal.pone.0111066>

```{r load-models}
model_names <- c(
  "Nakashima_2015_valproic_acid",
  "Nakashima_2015_valproic_acid_le18y",
  "Nakashima_2015_valproic_acid_ge19y"
)

# Resolve each model to an rxUi exactly once. readModelDb() returns the model
# FUNCTION, so `rxode2::rxode()` (not `attr()` / `environment()`) is what
# yields an object whose $population / $reference / $description work.
uis <- lapply(model_names, function(n) rxode2::rxode(readModelDb(n)))
names(uis) <- model_names

cat(uis[["Nakashima_2015_valproic_acid"]]$reference)
```

## Expression of Concern

**The source article carries a PLOS ONE Expression of Concern (January 2023,
[doi:10.1371/journal.pone.0279487](https://doi.org/10.1371/journal.pone.0279487)).**
Its stated grounds are **funding disclosure only**: the study was part-funded by
the Smoking Research Foundation, which has received financial support from the
tobacco industry, and PLOS ONE has prohibited tobacco-industry funding since
2010. The notice raises **no concern about the data, the methods, or the
results**, and it is **not a retraction**. The upstream population PK paper
(Ogusu 2014), from the same group and carrying the same funding
acknowledgement, has no such notice.

The Expression of Concern is recorded in each model file's `reference` metadata
so that it travels with the model, and is repeated in the Errata section below.
Users should be aware of it when citing these models.

# Population

Seventy-seven Japanese patients with epilepsy on maintenance sustained-release
valproic acid, treated at Kumamoto Saishunso National Hospital between June 1989
and April 2011, contributing 729 VPA concentration measurement points and one
binary efficacy outcome each (Nakashima 2015 Table 1). Mean age
15.2 +/- 8.2 years (range 0.8-36.9); 29 of 77 female (37.7%). The cohort is
strongly partial-onset (67 of 77, 87.0%; generalized 10, 13.0%), half
symptomatic and half cryptogenic, and 76.6% have intellectual disability.
Idiopathic epilepsy and severe retardation were exclusion criteria.

Valproic acid dose 1120.0 +/- 592.5 mg/day (range 50-3200), with a
model-predicted steady-state trough of 69.3 +/- 19.9 ug/mL (range 11.8-130.1).
Concomitant antiepileptic drugs were common: carbamazepine 61.0%, clobazam
42.9%, phenobarbital 40.3%, phenytoin 28.6%, zonisamide 33.8%, clonazepam
27.3%, topiramate 13.0%, gabapentin 10.4%, lamotrigine 3.9%. *SCN1A* rs3812718
genotypes were G/G 11.7%, G/A 53.2%, A/A 35.1% (A-allele frequency 61.7%),
consistent with Hardy-Weinberg equilibrium.

The efficacy endpoint is an over-50% reduction in seizure frequency relative to
the pre-VPA baseline, assessed over a window at least five times each patient's
own seizure interval. The ROC curve for logit(Pr) had an area under the curve of
0.823 (95% CI 0.793-0.853), and logit(Pr) = 0.1 was selected as the optimal
cut-off (sensitivity 71.8%, specificity 80.4%).

The same information is available programmatically:

```{r population}
str(uis[["Nakashima_2015_valproic_acid"]]$population, max.level = 1)
```

# Source trace

The per-parameter origin is recorded as an in-file comment beside each `ini()`
entry in `inst/modeldb/specificDrugs/Nakashima_2015_valproic_acid*.R`. Collected
here for review.

## PK layer -- all values fixed, all from Ogusu 2014

Nakashima 2015 states that "the individual PK parameters, determined from our
previously reported population PK model [8], were fixed in the PK-PD analysis",
where [8] is Ogusu 2014. All three model files therefore share one fixed PK
layer, taken from the **upstream source**, not from Nakashima 2015's own
restatement of it (see Errata).

| Parameter | Value | Source location |
|---|---|---|
| `lka` | 0.109 1/h | Ogusu 2014 Eq 5; Table 2 "Ka" |
| `lvc` | 21.4 L at 1000 mg/day | Ogusu 2014 Eq 6; Table 2 "Vd/F" |
| `lcl` | 0.559 L/h at 1000 mg/day | Ogusu 2014 Eq 7; Table 2 "CL/F" |
| `ltlag` | 3.00 h | Ogusu 2014 Eq 8; Table 2 "ALAG (h) 3.00 (Fixed)" |
| `e_dose_vc` | 1.52 | Ogusu 2014 Eq 6; Table 2 "Dose on Vd/F" |
| `e_dose_cl` | 0.596 | Ogusu 2014 Eq 7; Table 2 "Dose on CL/F" |
| `e_sexf_cl` | 0.917 | Ogusu 2014 Eq 7; Table 2 "Gender on CL/F" |
| `e_cbz_cl` | 1.19 | Ogusu 2014 Eq 7; Table 2 "CBZ on CL/F" |
| `e_pb_cl` | 1.12 | Ogusu 2014 Eq 7; Table 2 "PB on CL/F" |
| `e_pht_cl` | 1.43 | Ogusu 2014 Eq 7; Table 2 "PHT on CL/F" |
| `e_clb_cl` | 0.906 | Ogusu 2014 Eq 7; Table 2 "CLB on CL/F" |
| `etalka` | 7.77e-7 | Ogusu 2014 Table 2 "omega^2 on Ka" |
| `etalvc` | 1.83e-7 | Ogusu 2014 Table 2 "omega^2 on Vd/F" |
| `etalcl` | 0.0587 | Ogusu 2014 Table 2 "omega^2 on CL/F" |
| `etaltlag` | 4.48e-9 | Ogusu 2014 Table 2 "omega^2 on ALAG" |
| ODE structure | n/a | Ogusu 2014 Methods: one compartment, first-order absorption, lag time |

## PD layer

| Parameter | Value | Source location |
|---|---|---|
| **Eq 6 (all ages)** | | |
| `logit_ref` | 6.09 | Nakashima 2015 Table 3 Intercept (Eq 6 prints 6.1) |
| `e_age_logit` | 0.98 per year | Nakashima 2015 Table 3 "Age (years)"; Eq 6's `(Age/10)` is falsified by Table 4 |
| `e_cbz_logit` | -1.75 | Nakashima 2015 Table 3 Intercept block "CBZ" |
| `e_czp_logit` | -1.18 | Nakashima 2015 Table 3 Intercept block "CZP" |
| `e_snp_scn1a_rs3812718_ga_logit` | -5.87 | Nakashima 2015 Table 3 Intercept block "SCN1A G/A genotype" |
| `e_snp_scn1a_rs3812718_aa_logit` | -4.88 | Nakashima 2015 Table 3 Intercept block "SCN1A A/A genotype" |
| `e_ctrough_logit` | -13.5 per 100 ug/mL | Nakashima 2015 Table 3 Slope (Eq 6 prints 13.3; Table 4 selects -13.5) |
| `e_partial_slope` | +2.41 | Nakashima 2015 Table 3 Slope block "Partial seizure" |
| `e_pht_slope` | -3.62 | Nakashima 2015 Table 3 Slope block "PHT" |
| `e_tpm_slope` | -1.73 | Nakashima 2015 Table 3 Slope block "TPM" |
| `e_snp_scn1a_rs3812718_ga_slope` | +10.1 | Nakashima 2015 Table 3 Slope block "SCN1A G/A genotype" |
| `e_snp_scn1a_rs3812718_aa_slope` | +9.48 | Nakashima 2015 Table 3 Slope block "SCN1A A/A genotype" |
| `etalogit_ref` | 11.3 | Nakashima 2015 Table 3 "Individual random effect" |
| **Eq 7 (<= 18 y)** | | |
| `logit_ref` / `e_cbz_logit` / `e_pb_logit` | 7.73 / -4.88 / -1.93 | Nakashima 2015 Eq 7 (no parameter table exists) |
| SCN1A G/A, A/A on intercept | -4.75 / -4.30 | Nakashima 2015 Eq 7 |
| `e_ctrough_logit` | -10.9 per 100 ug/mL | Nakashima 2015 Eq 7 slope group |
| `e_cbz_slope` / `e_pht_slope` | +4.73 / -3.86 | Nakashima 2015 Eq 7 slope group |
| SCN1A G/A, A/A on slope | +7.62 / +7.60 | Nakashima 2015 Eq 7 slope group |
| `etalogit_ref` | 11.5 | Nakashima 2015 Results, paragraph after Eq 8 |
| **Eq 8 (>= 19 y)** | | |
| `logit_ref` / `e_czp_logit` | 10.3 / -2.56 | Nakashima 2015 Eq 8 (no parameter table exists) |
| SCN1A G/A on intercept | -9.88 | Nakashima 2015 Eq 8 |
| `e_snp_scn1a_rs3812718_aa_slope` | -14.3 per 100 ug/mL | Nakashima 2015 Eq 8; the only concentration-dependent term as printed |
| `etalogit_ref` | 9.71 | Nakashima 2015 Results, paragraph after Eq 8 |
| Logistic transform | n/a | Nakashima 2015 Eq 3 (`Pr = e^logit / (1 + e^logit)`) |
| Linear-in-trough logit | n/a | Nakashima 2015 Eq 4 |

Equations 1-8 are typeset as page images in the PLOS PDF and are not recoverable
from the trimmed-text extraction; every equation above was read from the
rendered page and cross-checked against the PDF text layer.

# Virtual cohort

Original patient-level data are not publicly available (the Data Availability
statement says "All relevant data are within the paper"). The cohort below is a
virtual population whose covariate distributions follow Nakashima 2015 Table 1.

```{r cohort}
# set.seed() seeds R's RNG only. rxode2's simulation RNG is partitioned per
# solver thread, so a cohort drawn here is NOT reproducible across machines
# with different thread counts. Every assertion below is written to hold for
# any cohort the model can produce (see pattern 12 of
# references/known-vignette-failure-patterns.md).
set.seed(20150141)

n_cohort <- 200L   # per-arm cap; 200 is ample for this validation

# Nakashima 2015 Table 1 marginal distributions.
cohort <- tibble(
  id  = seq_len(n_cohort),
  AGE = pmin(pmax(rnorm(n_cohort, 15.2, 8.2), 0.8), 36.9),
  SEXF = rbinom(n_cohort, 1, 0.377),
  DOSE_VPA_MGD = pmin(pmax(rnorm(n_cohort, 1120.0, 592.5), 50), 3200),
  SEIZURE_LOCUS_PARTIAL = rbinom(n_cohort, 1, 0.870),
  CONMED_CBZ = rbinom(n_cohort, 1, 0.610),
  CONMED_CZP = rbinom(n_cohort, 1, 0.273),
  CONMED_CLB = rbinom(n_cohort, 1, 0.429),
  CONMED_PB  = rbinom(n_cohort, 1, 0.403),
  CONMED_PHT = rbinom(n_cohort, 1, 0.286),
  CONMED_TPM = rbinom(n_cohort, 1, 0.130)
)

# SCN1A rs3812718: G/G 11.7%, G/A 53.2%, A/A 35.1% (G/G is the reference, so
# both indicators are 0 for it).
geno <- sample(c("GG", "GA", "AA"), n_cohort, replace = TRUE,
               prob = c(0.117, 0.532, 0.351))
cohort$SNP_SCN1A_RS3812718_GA <- as.integer(geno == "GA")
cohort$SNP_SCN1A_RS3812718_AA <- as.integer(geno == "AA")

covariate_cols <- setdiff(names(cohort), "id")
```

The paper's PK layer is a steady-state model: patients had received
sustained-release VPA for at least three weeks. `DOSE_VPA_MGD` is the *daily*
dose and is the covariate driving CL/F and Vd/F, so the per-administration `amt`
is `DOSE_VPA_MGD` divided by the number of daily administrations. The "predicted
trough concentration of VPA" is then the concentration at the end of a dosing
interval at steady state.

**The paper never states the administration frequency it assumed**, and the
predicted trough depends on it, so `make_events()` below is parameterised by
`ndose` and Gate 2 reports the sensitivity rather than committing to one value.

```{r events-helper}
# Build a dose-to-steady-state event table with a single trough observation at
# t = n_days * 24, plus any extra observation times requested.
#
# Observation rows use cmt = "central", the ODE STATE name -- never "Cc" or
# "prob_seizure50", which are algebraic observables. rxode2 returns every
# algebraic observable as an output column at those rows anyway.
n_days <- 50L   # 105 / 41 / 20 half-lives at 400 / 1120 / 2400 mg/day
tau    <- 24    # one day

make_events <- function(subj, ndose = 1L, extra_obs_times = numeric(0)) {
  tau_i <- tau / ndose                       # inter-dose interval

  doses <- subj |>
    tidyr::crossing(time = seq(0, n_days * tau - tau_i, by = tau_i)) |>
    mutate(evid = 1L, cmt = "depot", amt = DOSE_VPA_MGD / ndose)

  obs <- subj |>
    tidyr::crossing(time = c(n_days * tau, extra_obs_times)) |>
    mutate(evid = 0L, cmt = "central", amt = NA_real_)

  bind_rows(doses, obs) |>
    arrange(id, time, desc(evid)) |>
    as.data.frame()
}
```

# Gate 1 -- reproduce Table 4's six published optimal trough concentrations

This is the strongest available check on the PD layer. Nakashima 2015 Table 4
prints the optimal trough concentration for six simulated patients: the trough
at which logit(Pr) reaches the ROC cut-off of 0.1. Reproducing all six pins the
intercept block, the slope block, the age scaling **and** the concentration
scaling simultaneously.

Crucially, the check is driven **through the compiled model**, not by re-typing
Eq 6 in this vignette. Because `logit_seizure50` is exactly linear in `Cc` by
construction, two solves at two different doses recover the model's own
intercept and slope on the logit scale exactly; the optimal trough then follows
by inversion. Re-typing the equation here would validate the transcription
against itself and could pass a dropped term.

```{r gate1-table4}
# Table 4: all six cases are "Generalized" seizure locus with phenytoin
# co-administration and no other AED (Nakashima 2015 Table 4).
table4 <- tibble::tribble(
  ~case,     ~AGE, ~geno, ~printed_ugml,
  "Case 1",     5, "GG",  63.5,
  "Case 2",     5, "GA",  71.3,
  "Case 3",     5, "AA",  78.5,
  "Case 4",    10, "GG",  92.0,
  "Case 5",    10, "GA", 140.9,
  "Case 6",    10, "AA", 142.4
) |>
  mutate(
    SEXF = 0L, SEIZURE_LOCUS_PARTIAL = 0L,          # "Generalized"
    CONMED_PHT = 1L,                                 # the only listed AED
    CONMED_CBZ = 0L, CONMED_CZP = 0L, CONMED_CLB = 0L,
    CONMED_PB = 0L, CONMED_TPM = 0L,
    SNP_SCN1A_RS3812718_GA = as.integer(geno == "GA"),
    SNP_SCN1A_RS3812718_AA = as.integer(geno == "AA")
  )

# Typical-value model: zero out the logit random effect so the line we recover
# is the population prediction.
mod6_typ <- readModelDb("Nakashima_2015_valproic_acid") |> rxode2::zeroRe()

# Two arbitrary distinct daily doses give two points on the model's own
# (Cc, logit) line for each covariate pattern.
probe_doses <- c(800, 1600)

probe <- table4 |>
  select(-printed_ugml, -geno) |>
  tidyr::crossing(DOSE_VPA_MGD = probe_doses) |>
  mutate(id = row_number())

probe_sim <- rxode2::rxSolve(
  mod6_typ,
  events = make_events(probe),
  keep   = c("case", "DOSE_VPA_MGD")
) |>
  as.data.frame() |>
  filter(time == n_days * tau)

# rxSolve omits `id` for a single subject; restore defensively.
if (is.null(probe_sim$id)) probe_sim$id <- 1L
stopifnot(nrow(probe_sim) == nrow(probe))

optimal <- probe_sim |>
  group_by(case) |>
  arrange(DOSE_VPA_MGD, .by_group = TRUE) |>
  summarise(
    # exact, because logit_seizure50 is linear in Cc
    slope_per_ugml = diff(logit_seizure50) / diff(Cc),
    intercept      = first(logit_seizure50) - slope_per_ugml * first(Cc),
    .groups = "drop"
  ) |>
  # ROC cut-off logit(Pr) = 0.1 (Nakashima 2015 Results, Fig 2)
  mutate(model_ugml = (0.1 - intercept) / slope_per_ugml) |>
  inner_join(table4 |> select(case, AGE, geno, printed_ugml), by = "case") |>
  mutate(pct_diff = 100 * (model_ugml - printed_ugml) / printed_ugml) |>
  select(case, AGE, geno, printed_ugml, model_ugml, pct_diff)

# The gate must have had rows to test (see pattern 10).
stopifnot(nrow(optimal) == 6L, !anyNA(optimal$model_ugml))

optimal |>
  mutate(across(c(printed_ugml, model_ugml), \(x) round(x, 1)),
         pct_diff = round(pct_diff, 2)) |>
  dplyr::rename(
    "Case" = case, "Age (y)" = AGE, "SCN1A" = geno,
    "Published optimal trough (ug/mL)" = printed_ugml,
    "Model optimal trough (ug/mL)"     = model_ugml,
    "Difference (%)"                   = pct_diff
  ) |>
  knitr::kable(
    caption = paste(
      "Gate 1. Reproduces Table 4 of Nakashima 2015: the trough concentration",
      "at which logit(Pr) reaches the ROC cut-off of 0.1, for six simulated",
      "patients with generalized seizures on phenytoin."
    ),
    align = c("l", "r", "l", "r", "r", "r")
  )
```

```{r gate1-assert}
# DETERMINISTIC check: both sides are typical-value quantities, and the only
# error source is the 3-significant-figure rounding of the published Table 3
# estimates and Table 4 troughs. A tight bound is therefore correct here (this
# is the "same drawn parameters" case, not a cohort statistic -- contrast the
# cohort gates below). Realised max |difference| is 0.30%.
stopifnot(max(abs(optimal$pct_diff)) < 1)
```

All six published troughs are reproduced to within 0.3%. Three readings of
Eq 6 are settled by this, each independently confirmed by all six cases:

1. **Age enters per YEAR, not per decade.** Eq 6 prints `(Age/10) x 1.0`, but
   Table 3's row is labelled "Age (years)" with estimate 0.98. Using 0.98 per
   year reproduces Table 4 to 0.30%; using 0.98 per decade misses by 88.9%,
   and no concentration rescaling rescues it. The printed `/10` is a
   typesetting artefact.
2. **The base slope is Table 3's -13.5, not Eq 6's -13.3.** With -13.3 the
   maximum error is 3.24% rather than 0.29%.
3. **The published slope coefficients are per 100 ug/mL.** Back-solving the
   concentration scale from each of the six cases returns 0.01002 +/- 0.00001,
   i.e. exactly 1/100 within the printed rounding. Without it the logit would
   swing by roughly 1600 across the cohort's observed 11.8-130.1 ug/mL trough
   range.

# Gate 2 -- reproduce Table 1's predicted-trough distribution

Table 1 reports a model-predicted steady-state trough of 69.3 +/- 19.9 ug/mL
[11.8-130.1] at daily doses of 1120.0 +/- 592.5 mg [50-3200]. That gates the
**PK layer**, including the choice of the upstream Ogusu 2014 constants over
Nakashima 2015's restatement of them.

```{r gate2-trough}
# Solve the same cohort under once-daily and twice-daily administration of the
# same TOTAL daily dose, so the frequency assumption is visible rather than
# buried.
solve_cohort <- function(ndose) {
  ev <- make_events(cohort, ndose = ndose)
  stopifnot(!anyDuplicated(unique(ev[, c("id", "time", "evid")])))
  out <- rxode2::rxSolve(
    mod6_typ,                    # typical-value PK: the paper's "predicted" trough
    events = ev,
    keep   = covariate_cols
  ) |>
    as.data.frame() |>
    filter(time == n_days * tau)
  if (is.null(out$id)) out$id <- 1L
  stopifnot(nrow(out) == n_cohort, all(out$Cc > 0))
  out
}

sim_cohort    <- solve_cohort(1L)   # q24h; also reused by the Figure 1 VPC below
sim_cohort_q12 <- solve_cohort(2L)  # q12h

cohort_events <- make_events(cohort, ndose = 1L)

fmt <- function(x, d = 1) sprintf(paste0("%.", d, "f +/- %.", d, "f [%.1f-%.1f]"),
                                  mean(x), sd(x), min(x), max(x))

trough_summary <- tibble(
  Quantity  = c("Daily VPA dose (mg/day)",
                "Predicted trough (ug/mL), dosed q24h",
                "Predicted trough (ug/mL), dosed q12h"),
  Published = c("1120.0 +/- 592.5 [50-3200]",
                "69.3 +/- 19.9 [11.8-130.1]",
                "69.3 +/- 19.9 [11.8-130.1]"),
  Simulated = c(fmt(sim_cohort$DOSE_VPA_MGD),
                fmt(sim_cohort$Cc),
                fmt(sim_cohort_q12$Cc)),
  Difference = c(
    sprintf("%+.1f%% on the mean", 100 * (mean(sim_cohort$DOSE_VPA_MGD) - 1120) / 1120),
    sprintf("%+.1f%% on the mean", 100 * (mean(sim_cohort$Cc) - 69.3) / 69.3),
    sprintf("%+.1f%% on the mean", 100 * (mean(sim_cohort_q12$Cc) - 69.3) / 69.3)
  )
)

knitr::kable(
  trough_summary,
  caption = paste(
    "Gate 2. Simulated versus published predicted steady-state trough VPA",
    "(Nakashima 2015 Table 1). Typical-value PK over the published dose and",
    "covariate distributions, at two administration frequencies."
  ),
  align = c("l", "l", "l", "r")
)
```

```{r gate2-assert}
# COHORT statistic -> assert on the CENTRE with headroom, never on the extremes,
# which are one draw of a random cohort (pattern 12).
#
# Choosing the bound. The published mean is 69.3 ug/mL. Large-sample values for
# this model over these covariate distributions are -15.8% (q24h), -8.2% (q12h)
# and -7.3% (q8h); a 200-subject draw scatters a few points either side of
# those. 25% therefore admits every frequency assumption and the cohort noise,
# while still going red hard on a real transcription error: substituting the
# Nakashima Eqs 1-2 volume coefficient (110 L for 21.4 L) or dropping the dose
# exponents moves the mean by far more than that, and a unit error moves it by
# an order of magnitude. The residual gap is discussed below -- it is NOT
# tuned away.
for (nm in c("q24h", "q12h")) {
  x <- if (nm == "q24h") sim_cohort$Cc else sim_cohort_q12$Cc
  pct <- 100 * (mean(x) - 69.3) / 69.3
  cat(sprintf("%s: mean simulated trough %.1f ug/mL vs published 69.3 (%+.1f%%)\n",
              nm, mean(x), pct))
  stopifnot(abs(pct) < 25)
}

# The published range is 11.8-130.1 ug/mL. Assert the bulk of the cohort lands
# in a plausible envelope rather than asserting the min / max themselves.
stopifnot(
  median(sim_cohort$Cc) > 25,
  median(sim_cohort$Cc) < 120
)
```

The PK layer lands within roughly 8-16% of the published mean trough depending
on the assumed administration frequency, in both cases **below** it. This is
close enough to confirm the structural choice -- the Ogusu 2014 constants rather
than Nakashima 2015's restatement of them (see Errata), which differ by a factor
of five on the volume coefficient -- but it is not an exact reproduction, and the
residual gap is worth naming rather than tuning away. Three contributors, in
what we judge to be decreasing order of importance:

1. **Covariates are drawn independently here, but are correlated in the real
   cohort.** 61% of these patients take carbamazepine, 40% phenobarbital and
   29% phenytoin, and in this model those raise VPA clearance by 19%, 12% and
   43% respectively; over the published prevalences the cohort's mean clearance
   is about 22% higher than a comedication-free reference. Clinically, that
   induction is *compensated by prescribing a higher dose* -- but independent
   draws break the dose-comedication correlation, so the simulated cohort takes
   induction without the compensating dose. Scaling each subject's dose by their
   own clearance multiplier (a crude stand-in for full compensation) moves the
   q12h mean from -8.2% to -2.6%.
2. **The administration frequency is unknown** and is worth 8 percentage points
   on its own between q24h and q12h.
3. **The paper's predicted troughs are individual (empirical-Bayes) predictions,
   not typical-value ones.** With log-normal clearance the mean of the resulting
   trough distribution sits above the typical-value trough by roughly
   `exp(omega^2 / 2)`, which for the published `omega^2 = 0.0587` is about +3%.

None of these is a transcription question, so none is resolvable from the paper;
they are recorded in Assumptions and deviations below.

# Gate 3 -- PKNCA exposure characterisation and an exact internal identity

Nakashima 2015 reports **no** NCA parameters -- no Cmax, Tmax, AUC or
half-life -- because the analysis is an exposure-response model on a trough
metric, so there is nothing to compare a simulated NCA table against. What *is*
available is an exact internal identity that the PK layer must satisfy: at
steady state, the AUC over one dosing interval equals dose divided by apparent
clearance. With the random effects zeroed there is no residual error and no IIV,
so this is a deterministic dimensional check on the whole PK chain (dose units,
volume units, the dose-scaling exponents, and the concentration definition
`Cc = central / vc`).

Following the standing guidance, the NCA is run on the **typical-value** solve
rather than over a full-IIV cohort, and on a densely sampled final dosing
interval.

```{r gate3-pknca}
# Three representative dose levels spanning the published range.
nca_subj <- cohort |>
  slice(1:3) |>
  mutate(
    DOSE_VPA_MGD = c(400, 1120, 2400),
    regimen = paste0(DOSE_VPA_MGD, " mg/day"),
    # Neutral covariate pattern so the identity is easy to state.
    SEXF = 0L, CONMED_CBZ = 0L, CONMED_PB = 0L, CONMED_PHT = 0L,
    CONMED_CLB = 0L, CONMED_CZP = 0L, CONMED_TPM = 0L,
    SEIZURE_LOCUS_PARTIAL = 1L, AGE = 15.2,
    SNP_SCN1A_RS3812718_GA = 0L, SNP_SCN1A_RS3812718_AA = 0L
  )

# Dense grid over the FINAL dosing interval, expressed as absolute times.
last_start <- (n_days - 1L) * tau
nca_events <- make_events(
  nca_subj,
  extra_obs_times = last_start + seq(0, tau, by = 0.25)
)

sim_nca_raw <- rxode2::rxSolve(
  mod6_typ, events = nca_events,
  keep = c("regimen", "DOSE_VPA_MGD")
) |>
  as.data.frame()
if (is.null(sim_nca_raw$id)) sim_nca_raw$id <- 1L

# Re-base time to the start of the final interval so PKNCA sees a clean
# single-interval profile anchored at time 0.
sim_nca <- sim_nca_raw |>
  filter(time >= last_start, !is.na(Cc)) |>
  mutate(time = time - last_start) |>
  select(id, time, Cc, regimen, DOSE_VPA_MGD) |>
  arrange(id, regimen, time)

# Filter is `!is.na(Cc)` ONLY -- adding `time > 0` or `Cc > 0` would drop the
# time-zero row PKNCA needs to anchor AUC.
stopifnot(all(sim_nca$Cc >= 0), any(sim_nca$time == 0))

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | regimen + id)

dose_df <- sim_nca |>
  distinct(id, regimen, DOSE_VPA_MGD) |>
  mutate(time = 0, amt = DOSE_VPA_MGD)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | regimen + id)

intervals <- data.frame(
  start = 0, end = tau,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)

nca_tab <- as.data.frame(nca_res) |>
  select(regimen, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

# Exact identity: AUC(0-tau, ss) == Dose / (CL/F).
cl_by_regimen <- sim_nca_raw |>
  filter(time == n_days * tau) |>
  distinct(regimen, DOSE_VPA_MGD, cl)

# The PD driver is the PRE-DOSE trough, i.e. the concentration at the start
# (equivalently, at steady state, the end) of the interval. That is NOT the
# same as PKNCA's `cmin`: with a 3 h absorption lag the profile DECLINES from
# the dose time until absorption begins, so the within-interval minimum sits
# at t = 3 h. Both are reported below; only the end-of-interval value is the
# paper's "trough".
interval_ends <- sim_nca |>
  group_by(regimen) |>
  summarise(
    c_predose = Cc[which.min(time)],
    c_end     = Cc[which.max(time)],
    .groups   = "drop"
  )

identity_tab <- nca_tab |>
  inner_join(cl_by_regimen, by = "regimen") |>
  inner_join(interval_ends, by = "regimen") |>
  mutate(
    auc_expected = DOSE_VPA_MGD / cl,
    auc_pct_diff = 100 * (auclast - auc_expected) / auc_expected,
    # Steady-state periodicity: C(tau) must equal C(0).
    periodicity_pct = 100 * abs(c_end - c_predose) / c_predose
  )

stopifnot(nrow(identity_tab) == 3L, !anyNA(identity_tab$auc_pct_diff))

identity_tab |>
  transmute(
    regimen,
    cmax = round(cmax, 1), tmax = round(tmax, 2),
    cmin = round(cmin, 1), c_end = round(c_end, 1),
    half.life = round(half.life, 1),
    auclast = round(auclast, 1),
    auc_expected = round(auc_expected, 1),
    auc_pct_diff = round(auc_pct_diff, 3)
  ) |>
  dplyr::rename(
    "Regimen"                       = regimen,
    "Cmax (ug/mL)"                  = cmax,
    "Tmax (h)"                      = tmax,
    "Cmin (ug/mL, at end of lag)"   = cmin,
    "Ctrough = C(tau) (ug/mL)"      = c_end,
    "t1/2 (h)"                      = half.life,
    "AUC0-24,ss (ug*h/mL)"          = auclast,
    "Dose / (CL/F) (ug*h/mL)"       = auc_expected,
    "Difference (%)"                = auc_pct_diff
  ) |>
  knitr::kable(
    caption = paste(
      "Gate 3. PKNCA exposure metrics over the final steady-state dosing",
      "interval (typical-value solve), and the exact mass-balance identity",
      "AUC(0-24,ss) = Dose / (CL/F). Nakashima 2015 publishes no NCA table,",
      "so the identity is the available check. Note Cmin and Ctrough differ:",
      "see the text."
    ),
    align = c("l", rep("r", 8))
  )
```

```{r gate3-assert}
# All three checks below are DETERMINISTIC (typical-value solve, no IIV, no
# residual error), so tight bounds are correct -- contrast Gate 2, which is a
# cohort statistic and is bounded loosely.

# (a) Mass balance. Residual difference is trapezoidal-integration error on a
# 0.25 h grid plus the finite approach to steady state over 50 days.
stopifnot(max(abs(identity_tab$auc_pct_diff)) < 1.5)

# (b) Steady state actually reached: C(tau) == C(0). The residual is the
# finite approach to steady state, which is first-order in the number of
# half-lives elapsed. 50 days of q24h dosing is 105 / 41 / 20 half-lives at
# 400 / 1120 / 2400 mg/day (the half-life rises steeply with dose because the
# volume exponent is 1.52), and the realised periodicity 4.9e-13 / 4.5e-11 /
# 2.9e-5 % tracks the predicted 100*2^-n_halflives of 1.8e-30 / 5.5e-11 /
# 8.7e-5 % closely. Bound set an order of magnitude above the slowest case;
# it still goes red hard if the dosing run is shortened (at 10 days the
# 2400 mg/day arm sits near 6%).
stopifnot(max(identity_tab$periodicity_pct) < 1e-3)

# (c) The PD layer's trough is the END-of-interval concentration read directly
# off the solve -- confirm Gate 1 / Gate 2 and this NCA agree on what "trough"
# means.
trough_direct <- sim_nca_raw |>
  filter(time == n_days * tau) |>
  select(regimen, Cc_direct = Cc)
chk <- identity_tab |> inner_join(trough_direct, by = "regimen")
stopifnot(nrow(chk) == 3L,
          max(abs(100 * (chk$c_end - chk$Cc_direct) / chk$Cc_direct)) < 1e-6)

# (d) Cmin is strictly BELOW Ctrough, because the 3 h absorption lag makes the
# profile fall before it rises. Deterministic, so the ordering is safe to
# assert; it is recorded because conflating the two is an easy error.
stopifnot(all(chk$cmin < chk$c_end))
```

The mass-balance identity holds to better than 1.5% at every dose level, and the
steady-state periodicity condition `C(tau) = C(0)` holds to between 5e-13% and
3e-5% depending on dose, so 50 days of once-daily dosing has genuinely reached
steady state everywhere in the range.

**Cmin is not the trough here, and the difference is mechanistic.** Because the
absorption lag is 3 h, the concentration *declines* from each dose time until
absorption begins, so the within-interval minimum occurs at exactly t = 3 h at
every dose level -- 2.6% to 11.9% below the pre-dose value across the three
regimens (at 1120 mg/day: 69.6 ug/mL at t = 3 h versus 73.3 ug/mL pre-dose).
The paper's "predicted trough concentration of VPA" is the pre-dose
concentration -- equivalently, at steady state, the value at the end of the
interval -- and that is what drives the PD layer. Any downstream analysis that
substituted an NCA `cmin` for the trough would bias the exposure metric low by
up to ~12% and shift the whole exposure-response relationship. This is a
general trap for lagged extravascular models: the usual shortcut that an oral
`Cmin` over an interval is its end-of-interval value holds only when there is
**no** absorption lag.

**The tabulated `half.life` is a within-interval estimate, not the model's
terminal half-life.** The model's true terminal half-life is
`log(2) / (CL/F / (Vd/F))`, which is 11.4 / 29.5 / 59.6 h at 400 / 1120 / 2400
mg/day, whereas PKNCA reports 17.8 / 43.8 / 86.9 h -- about 1.5-fold longer.
The cause is the slow absorption (`Ka = 0.109 1/h`, an absorption half-life of
6.4 h) combined with the 3 h lag: within a single steady-state interval, drug is
still being absorbed throughout the apparent decline, which flattens the
log-linear slope PKNCA regresses on. This is a window artefact, not a model
defect; the terminal slope is only recoverable from a washout profile, which a
steady-state interval does not contain.

Note also that the apparent volume carries a steep dose exponent (1.52), so
Vd/F ranges from a fraction of a litre at 50 mg/day to well over 100 L at
3200 mg/day, and the terminal half-life varies more than five-fold across the
cohort's dose range as a result. That is Ogusu 2014's published
parameterisation and is reproduced faithfully, but it means the PK layer should
not be extrapolated outside the 50-3200 mg/day range the source cohort spans.

# Replicate Figure 1 -- the exposure-response direction

Figure 1 is a visual predictive check of the observed proportion of patients
with an over-50% seizure-frequency reduction against the predicted trough. The
observed proportion falls from roughly 0.50 near 45 ug/mL to roughly 0.29 near
100 ug/mL, i.e. higher predicted troughs are associated with *worse* seizure
control in this cohort -- the paper's central and initially counter-intuitive
finding, which is what motivates an *optimal* (upper-limit) trough rather than
a "higher is better" target.

```{r figure-1}
# Full-IIV cohort: the logit random effect (variance 11.3) is what makes this a
# VPC rather than a typical-value curve.
mod6 <- readModelDb("Nakashima_2015_valproic_acid")

sim_vpc <- rxode2::rxSolve(
  mod6, events = cohort_events, keep = covariate_cols
) |>
  as.data.frame() |>
  filter(time == n_days * tau)
if (is.null(sim_vpc$id)) sim_vpc$id <- 1L
stopifnot(nrow(sim_vpc) == n_cohort)

binned <- sim_vpc |>
  mutate(trough_bin = cut(Cc, breaks = c(0, 40, 60, 80, 100, Inf),
                          labels = c("<40", "40-60", "60-80", "80-100", ">100"))) |>
  group_by(trough_bin) |>
  summarise(n = n(), mean_prob = mean(prob_seizure50),
            median_trough = median(Cc), .groups = "drop") |>
  filter(n >= 5)

ggplot(binned, aes(median_trough, mean_prob)) +
  geom_line() +
  geom_point(aes(size = n)) +
  geom_hline(yintercept = c(0.29, 0.50), linetype = "dashed", colour = "grey50") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Predicted steady-state trough VPA (ug/mL)",
    y = "Mean model-predicted P(>50% seizure reduction)",
    size = "Subjects",
    title = "Figure 1 - exposure-response direction",
    caption = paste(
      "Replicates the direction of Figure 1 of Nakashima 2015. Dashed lines",
      "mark the paper's observed proportions of ~0.50 near 45 ug/mL and",
      "~0.29 near 100 ug/mL."
    )
  )
```

```{r figure-1-assert}
# TREND, not step-by-step monotonicity, and no assertion on the sign of a
# near-zero quantity (pattern 12). The cohort-average slope is genuinely
# negative and not near zero: with SCN1A G/A and A/A carrying large POSITIVE
# slope offsets (+10.1, +9.48) against a -13.5 base, the cohort-mean slope is
# roughly -3 per 100 ug/mL, so the fall across the observed trough range is
# real but modest.
stopifnot(nrow(binned) >= 3L)
stopifnot(binned$mean_prob[nrow(binned)] < binned$mean_prob[1])

# Probabilities must be a valid probability, which also guards the expit().
stopifnot(all(sim_vpc$prob_seizure50 >= 0), all(sim_vpc$prob_seizure50 <= 1))
```

# Age-subgroup models (Eqs 7 and 8)

The authors also fit the two age subgroups separately and note in the
Discussion that "the extent of the impact of the *SCN1A* genotype and the
co-administrated AEDs on the logit(Pr) values differed between these models".
The comparison below shows the three models side by side over a common trough
range, holding covariates at a neutral reference pattern (no concomitant AED,
generalized locus) and varying genotype.

```{r subgroup-comparison}
subgroup_grid <- tidyr::crossing(
  model = model_names,
  geno  = c("GG", "GA", "AA"),
  DOSE_VPA_MGD = seq(200, 3200, by = 200)
) |>
  mutate(
    SEXF = 0L, AGE = 15.2, SEIZURE_LOCUS_PARTIAL = 0L,
    CONMED_CBZ = 0L, CONMED_CZP = 0L, CONMED_CLB = 0L,
    CONMED_PB = 0L, CONMED_PHT = 0L, CONMED_TPM = 0L,
    SNP_SCN1A_RS3812718_GA = as.integer(geno == "GA"),
    SNP_SCN1A_RS3812718_AA = as.integer(geno == "AA")
  )

subgroup_sim <- lapply(model_names, function(nm) {
  subj <- subgroup_grid |>
    filter(model == nm) |>
    mutate(id = row_number())
  out <- rxode2::rxSolve(
    readModelDb(nm) |> rxode2::zeroRe(),
    events = make_events(subj),
    keep   = c("geno", "DOSE_VPA_MGD")
  ) |>
    as.data.frame() |>
    filter(time == n_days * tau)
  if (is.null(out$id)) out$id <- 1L
  out$model <- nm
  out
}) |>
  bind_rows()

stopifnot(nrow(subgroup_sim) == nrow(subgroup_grid))

subgroup_sim |>
  mutate(model = factor(model, levels = model_names,
                        labels = c("Eq 6 (all ages)", "Eq 7 (<=18 y)",
                                   "Eq 8 (>=19 y)"))) |>
  ggplot(aes(Cc, prob_seizure50, colour = geno)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~model) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Predicted steady-state trough VPA (ug/mL)",
    y = "P(>50% seizure reduction)", colour = "SCN1A\nrs3812718",
    title = "Exposure-response by model and SCN1A rs3812718 genotype",
    caption = paste(
      "Neutral covariate pattern (no concomitant AED, generalized locus,",
      "age 15.2 y where age enters). Eq 8's flat G/G and G/A curves are the",
      "printed-equation ambiguity described in the Errata."
    )
  )
```

```{r subgroup-assert}
# Eq 8 as printed makes only the A/A group concentration-dependent. That is a
# DETERMINISTIC structural consequence of the encoding, so it is asserted
# exactly -- and it is the single most important thing for a user of the
# >=19 y model to know.
eq8 <- subgroup_sim |> filter(model == "Nakashima_2015_valproic_acid_ge19y")
flat <- eq8 |> filter(geno %in% c("GG", "GA")) |>
  group_by(geno) |> summarise(spread = diff(range(prob_seizure50)), .groups = "drop")
stopifnot(nrow(flat) == 2L, all(flat$spread < 1e-8))

# The A/A group under Eq 8 is not flat, and decreases with trough.
aa <- eq8 |> filter(geno == "AA") |> arrange(Cc)
stopifnot(diff(range(aa$prob_seizure50)) > 0.05,
          aa$prob_seizure50[nrow(aa)] < aa$prob_seizure50[1])
```

Eq 8's G/G and G/A curves are exactly flat because, as printed, Eq 8 carries no
parentheses and only its A/A term multiplies the trough concentration. This is
the paper as published, not an encoding slip; see the Errata.

# Assumptions and deviations

- **Dosing frequency is unknown and is not resolvable from the paper.** The
  paper reports a *daily* dose and a steady-state trough but never states the
  administration frequency assumed when predicting troughs. Gate 2 reports both
  q24h and q12h rather than selecting one; the figures and the subgroup
  comparison use q24h as the simplest reading. Do **not** read the choice as
  validated: it is worth about 8 percentage points on the predicted mean trough,
  and neither frequency reproduces the published 69.3 ug/mL exactly.
- **Covariate distributions are drawn independently** from the Table 1
  marginals. The paper publishes no correlation structure among age, dose,
  concomitant AEDs and genotype. This is the largest single contributor to
  Gate 2's residual gap: enzyme-inducing comedications raise clearance (by 19%
  for carbamazepine, 12% for phenobarbital and 43% for phenytoin in this model)
  and are clinically compensated with a higher dose, but independent draws take
  the induction without the compensation. Table 1 also reports demographics for
  the **full** cohort only, never broken out by age subgroup, so the Eq 7 and
  Eq 8 comparisons above use the same marginals.
- **Predicted troughs here are typical-value, not empirical-Bayes.** The paper's
  Table 1 trough distribution comes from individual PK predictions; with
  log-normal clearance that mean sits roughly `exp(omega^2/2)` (about +3%) above
  the typical-value trough.
- **Between-subject variability is the paper's logit random effect only.** The
  variance is read as being on the logit scale: Table 3 gives 11.3 and the
  Results state the base model was 12.9 and the final model 11.3, "a 12.4%
  decrease" -- and (12.9 - 11.3)/12.9 = 12.4% confirms the two figures are the
  same quantity on the same scale.
- **No residual error is published or estimable.** The source likelihood is
  Bernoulli on a binary per-patient outcome (NONMEM LAPLACIAN), so there is no
  sigma to transcribe. Each model file attaches a small fixed additive
  placeholder (`addSd_prob_seizure50 = 0.001`) purely so the observation
  equation is valid for simulation; **this value is not from the paper**. The
  upstream PK residual error (Ogusu 2014 Table 2, sigma^2 = 0.0617
  proportional, about 24.8% CV) is deliberately *not* carried, because the PK
  parameters were fixed into the PD fit and so the PK residual does not enter
  this model's likelihood.
- **The PK layer is not paper-derived from Nakashima 2015** but from the
  upstream Ogusu 2014 publication that Nakashima 2015 cites and fixes its PK
  from. See the Errata for the discrepancy this resolves.
- **The PK layer should not be extrapolated outside 50-3200 mg/day.** The
  published dose exponent on apparent volume is 1.52, which makes Vd/F -- and
  therefore the terminal half-life -- vary steeply and non-physiologically with
  dose across the range.

# Errata and source defects

Four defects in the printed source were found and resolved. All are recorded in
the model files as well.

## 1. Nakashima 2015 misquotes the upstream PK model on all nine constants

Nakashima 2015 restates the inherited Ogusu 2014 PK model in its own Eqs 1-2,
and disagrees with the source on every constant:

| Constant | Ogusu 2014 (source, used here) | Nakashima 2015 (restatement) |
|---|---|---|
| Vd/F coefficient | **21.4 L** | 110 L |
| Vd/F dose exponent | 1.52 | 1.51 |
| CL/F coefficient | 0.559 L/h | 0.577 L/h |
| CL/F dose exponent | 0.596 | 0.535 |
| female on CL/F | 0.917 | 0.875 |
| CBZ on CL/F | 1.19 | 1.22 |
| PB on CL/F | 1.12 | 1.10 |
| PHT on CL/F | 1.43 | 1.40 |
| CLB on CL/F | 0.906 | 0.915 |

These are not rounding differences, and they are not Ogusu 2014's bootstrap
medians either. Ogusu 2014's printed Eqs 5-8 and its Table 2 agree with each
other exactly, so the source is internally consistent and the restatement
matches neither. The decisive discriminator is physiological: the restated
`Vd/F = 110 x (Dose/1000)^1.51` gives 110-130 L at this cohort's doses, about
2.8-3.3 L/kg for a cohort whose mean age is 15 years, which is impossible for
valproic acid -- a highly protein-bound drug with an apparent volume of roughly
0.1-0.4 L/kg. Ogusu 2014's 21.4 L is about 0.6 L/kg.

**Resolution (operator ruling, sidecar `oare_PMC9833507` q2):** encode the
upstream Ogusu 2014 values, since the cited source has authority over a
restatement of it, and document the discrepancy here. Gate 2 above confirms the
choice reproduces Nakashima 2015's own published trough distribution.

## 2. Eq 6's age term prints a spurious `/10`

Eq 6 prints `(Age/10) x 1.0` while Table 3 labels the row "Age (years)" with
estimate 0.98. Inverting Eq 6 against Table 4's six published optimal troughs
(Gate 1) reproduces them to 0.30% with 0.98 **per year** and misses by 88.9%
with the per-decade reading. The `/10` is a typesetting artefact; the per-year
reading is used.

## 3. Eq 6's base slope disagrees with Table 3, and Table 3 has its own sign typo

Eq 6 prints a base slope of 13.3 where Table 3 reports -13.5. Table 3
reproduces Table 4 to 0.29% and Eq 6's value to 3.24%, so Table 3's estimate is
used (it also carries more significant figures). Separately, Table 3's
*bootstrap median* for the Intercept is printed as -5.66 while its own 95%
confidence interval is 0.18 to 13.80 -- the median falls outside its own
interval, so the sign is wrong and it should read +5.66. This is not
load-bearing here because the NONMEM Estimate column (6.09) is what the model
uses.

## 4. Eq 8 has an unmatched parenthesis

Eqs 6 and 7 both read `... - ( base_slope + covariate terms ) x Ctrough`. Eq 8,
verified against both the rendered page image and the PDF text layer, carries
**no parentheses at all**:

```
Logit(Pr) = 10.3 - 2.56^CZP - 9.88^{SCN1A GA} - 14.3^{SCN1A AA}
          x Predicted trough concentration of VPA                        (8)
```

Read literally, only the A/A term multiplies the concentration, so G/G and G/A
adults receive a concentration-*independent* response probability. Two points
make this the encoded reading:

1. The literal reading and the "restore a dropped opening parenthesis" reading
   **coincide**. An opening parenthesis before `9.88` would require the inner
   operator to be `+`, because the group as a whole is preceded by `-` (that is
   exactly how Eqs 6 and 7 are typeset -- the interior signs are inverted
   relative to the expanded form). The printed operator before `9.88` is `-`,
   so no consistent parenthesisation places `9.88` inside a slope group.
2. There is **no anchor** against which to test Eq 8. Table 4's six simulated
   patients are aged 5 and 10, so they exercise Eq 6 only; Figure 1's visual
   predictive check is the all-age model; and Eq 8 has no parameter table. Eq 8
   also prints no base slope of its own, which is itself a consequence of the
   missing grouping.

**Resolution (operator ruling, sidecar `oare_PMC9833507` q4):** encode Eq 8
exactly as printed and document the ambiguity. A user who prefers the
alternative reading -- all three genotype terms inside a slope group -- must
supply the base slope themselves, as it is not recoverable from any source on
disk.

## Other notes on the source

- Table 1's row "Predictive trough concentration of VPA" gives its units as
  `mg/day`, which is a copy-paste error from the row above; Table 4 and the
  Figure 1 x-axis both give `ug/ml`.
- The covariate gloss printed after Eq 8 still defines "partial seizure" and
  lists "CBZ, PB, CZP or PHT" even though Eq 7 uses no partial-seizure term and
  Eq 8 uses only CZP. That gloss is shared boilerplate carried over from Eq 6
  and must not be read as evidence that a term was lost in typesetting.
- Ogusu 2014 is internally inconsistent about its residual error model: its
  Methods and its Table 2 row label both say *proportional*, while its Results
  text says *additive*. Proportional is adopted (2:1, and the Table 2 label is
  explicit) -- though as noted above the PK residual does not enter these
  models.
- Neither paper carries any other erratum or correction (EuropePMC
  `commentCorrectionList` checked for both).
```
