---
title: "Adalimumab in juvenile idiopathic arthritis (Nassar-Sheikh Rashid 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adalimumab in juvenile idiopathic arthritis (Nassar-Sheikh Rashid 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
# readModelDb() returns the model FUNCTION; rxode2::rxode() resolves it to the
# ui exactly once so every downstream accessor ($theta, $population, ...) works.
ui <- rxode2::rxode(readModelDb("NassarSheikhRashid_2024_adalimumab"))
```

* Citation: `r ui$reference`
* Article: <https://doi.org/10.1007/s40272-024-00629-7> (open access; PMC11192828)
* Upstream model: `modellib("Kang_2020_adalimumab_phase3_extension")`

This is the first population PK analysis of adalimumab in juvenile idiopathic
arthritis (JIA). The authors first showed that five published adalimumab models
-- one paediatric Crohn's disease model, one adult rheumatoid arthritis (RA)
model and three RA models from Kang 2020 -- all described their JIA
concentration-time data adequately. They then took the best-fitting structure,
Kang 2020 model 3 (the RA maintenance-phase model, carried in nlmixr2lib as
`Kang_2020_adalimumab_phase3_extension`), and re-estimated apparent clearance,
its covariates, the inter-individual variability on clearance and the residual
error on the JIA data. Everything else is fixed at Kang's values, because 78
concentrations from 50 patients cannot support a full re-fit.

The clinically interesting result is the covariate model on `CL/F`: alongside
the expected body-weight and anti-drug-antibody effects, adalimumab clearance is
28% *lower* on concomitant methotrexate and 44% *higher* in patients with
comorbid uveitis. The uveitis association is novel to this paper and offers a
pharmacokinetic rationale for the escalated dosing that JIA-with-uveitis
patients are observed to need.

# Population

```{r population}
p <- ui$population
tibble::tibble(
  Field = c("Species", "Subjects", "Observations", "Age", "Body weight",
            "Female", "Disease state", "Dosing", "Region"),
  Value = c(p$species, as.character(p$n_subjects), as.character(p$n_observations),
            p$age_median, paste0(p$weight_median, " (IQR ", "29.4-59.8 kg)"),
            paste0(p$sex_female_pct, "%"), p$disease_state, p$dose_range, p$regions)
) |>
  knitr::kable(caption = "Study population (Nassar-Sheikh Rashid 2024 Table 1 and Section 2.1).")
```

Fifty children treated with subcutaneous adalimumab at a single Dutch centre
contributed 78 therapeutic-drug-monitoring samples collected between January
2013 and August 2023 (Table 1, Section 2.1). Sampling was extremely sparse and
opportunistic: 33 of the 50 patients contributed only one sample, and samples
were drawn at arbitrary times within the dosing interval rather than to a PK
schedule. Median body weight was 49 kg (IQR 29.4-59.8), mean age 11.8 years
(SD 3.9), and 72% were female. Anti-drug antibodies were detected in 9 patients
(18%), 39 (78%) were on concomitant methotrexate, and 8 (16%) had active
uveitis. Median C-reactive protein was very low at 0.55 mg/L (IQR 0.3-1.2),
consistent with the low-grade inflammatory burden typical of JIA. Race and
ethnicity were not reported.

The same information is available programmatically via
`readModelDb("NassarSheikhRashid_2024_adalimumab")()$population`.

# Source trace

Every `ini()` entry carries an in-file comment naming its origin. The table
below collects them for review. Values marked *fixed (Kang)* were not estimated
on the JIA data; they are carried over from Kang 2020 model 3 and appear in
Table 2 flagged with an asterisk.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` (CL/F) | 0.374 L/day per 70 kg | Table 2, row `CL/F, L/day (WT/70)^0.75` (RSE 10.0%) |
| `lvc` (V1/F) | 2.67 L per 70 kg | Table 2, row `V1/F*` -- fixed (Kang) |
| `lq` (Q/F) | 1.7 L/day per 70 kg | Table 2, row `Q/F*` -- fixed (Kang) |
| `lvp` (V2/F) | 3.94 L per 70 kg | Table 2, row `V2/F*` -- fixed (Kang) |
| `lka` | 0.262 /day | Table 2, row `ka*` -- fixed (Kang) |
| `ld1` (zero-order input duration) | 0.121 day | Table 2, row `D1*` -- fixed (Kang) |
| `lfdepot` | 1 (structural anchor) | Table 2 reports CL/F, V1/F, Q/F, V2/F throughout; F not estimated |
| `e_wt_cl`, `e_wt_q` | 0.75 | Table 2 row labels `(WT/70)^0.75`; Section 3.3 "standard allometric scaling ... reference weight of 70 kg" |
| `e_wt_vc`, `e_wt_vp` | 1 | Table 2 row labels `(WT/70)^1`; Section 3.3 |
| `e_ada_pos_cl` | log(2.08) | Table 2, row `(theta_ADA)^ADA*` -- fixed; Section 3.3 derivation from Kang model 3 |
| `e_conmed_mtx_cl` | log(0.720) | Table 2, row `(theta_MTX)^MTX` (RSE 10.8%) |
| `e_crp_cl` | 0.165 | Table 2, row `(CRP/0.6)^theta_CRP` (RSE 31.6%) |
| `e_dis_uveitis_cl` | log(1.44) | Table 2, row `(theta_UV)^UVEITIS` (RSE 16.4%) |
| `etalcl` | 0.0754785 | Table 2, `IIV CL/F` = 28.0% CV; omega^2 = log(1 + CV^2) |
| `etalvc`, `etalq`, `etalvp`, `etalka` (diagonal) | 0.585, 1.31, 0.150, 0.439 | Table 2, `IIV V1/F* 89.2%`, `Q/F* 165%`, `V2/F* 40.3%`, `ka* 74.2%` -- fixed (Kang) |
| omega off-diagonals of the fixed BLOCK(4) | -0.205, -0.138, 0.287, 0.0438, 0.559, 0.119 | **Not in this paper.** Section 3.3 "keeping the full omega variance-covariance matrix for the other parameters as described previously"; values read from Kang 2020 Table 4 via `modellib("Kang_2020_adalimumab_phase3_extension")` |
| `addSd` | 3.61 mg/L | Table 2, `Additive, mg/L` = 13.0 with "(SD 3.61 mg/L)" alongside |
| Covariate functional forms | n/a | Equation 1 (categorical power `theta_p * theta_cov^COV`) and Equation 2 (continuous power `theta_p * (COV/refCOV)^theta_cov`) |
| Sequential zero- then first-order SC absorption | n/a | Section 3.3, carried over from Kang 2020 model 3 (`dur(depot) <- d1`, then first-order `ka`) |
| ODE structure (2 compartments, linear elimination) | n/a | Section 2.3 and 3.2 (Kang model 3 structure) |

# Covariate model: reproducing the paper's own arithmetic

Section 3.3 states four quantitative claims about the covariate model. Each is a
deterministic consequence of the `ini()` values, so these are exact checks: they
compare the packaged model's `cl` against the number printed in the paper, with
no simulation noise on either side.

```{r covariate-claims}
REF_WT  <- 70
REF_CRP <- 0.6

# One typical (zero-eta) subject per covariate scenario.
cov_arms <- tibble::tibble(
  arm         = c("Reference (70 kg, ADA-, no MTX, no uveitis, CRP 0.6)",
                  "ADA-positive", "On methotrexate", "Active uveitis",
                  "CRP 10 mg/L", "35 kg, otherwise reference"),
  WT          = c(REF_WT, REF_WT, REF_WT, REF_WT, REF_WT, 35),
  ADA_POS     = c(0, 1, 0, 0, 0, 0),
  CONMED_MTX  = c(0, 0, 1, 0, 0, 0),
  DIS_UVEITIS = c(0, 0, 0, 1, 0, 0),
  CRP         = c(REF_CRP, REF_CRP, REF_CRP, REF_CRP, 10, REF_CRP)
) |>
  dplyr::mutate(id = dplyr::row_number())

ev_cov <- dplyr::bind_rows(
  cov_arms |> dplyr::transmute(id, time = 0, amt = 40, evid = 1L,
                               cmt = "depot", rate = -2),
  cov_arms |> dplyr::transmute(id, time = 1, amt = NA_real_, evid = 0L,
                               cmt = "central", rate = 0)
) |>
  dplyr::left_join(cov_arms, by = "id") |>
  dplyr::arrange(id, time, dplyr::desc(evid))

# omega = NA is mandatory: zeroRe() alone does not stop rxode2 re-using an
# omega left in the solve options by a previous stochastic solve.
sim_cov <- rxode2::rxSolve(
  rxode2::zeroRe(ui), events = ev_cov, omega = NA,
  keep = c("arm"), returnType = "data.frame"
)

cl_by_arm <- sim_cov |>
  dplyr::group_by(arm) |>
  dplyr::summarise(cl = unique(round(cl, 10)), .groups = "drop")
stopifnot(nrow(cl_by_arm) == nrow(cov_arms))   # one CL per arm, i.e. truly typical

cl_of <- function(a) cl_by_arm$cl[match(a, cl_by_arm$arm)]
cl_ref <- cl_of("Reference (70 kg, ADA-, no MTX, no uveitis, CRP 0.6)")

claims <- tibble::tribble(
  ~Claim,                                                            ~Paper,  ~Model,
  "Typical CL/F at 70 kg (Table 2, abstract '0.37 L/day per 70 kg')", 0.374,   cl_ref,
  "CL/F on methotrexate (Section 4: 'reduced by 28% to 0.27 L/day')", 0.269,   cl_of("On methotrexate"),
  "ADA-positive CL/F ratio (Section 3.3: '108% increase')",           2.08,    cl_of("ADA-positive") / cl_ref,
  "Uveitis CL/F ratio (Section 3.3: '44% higher clearance')",         1.44,    cl_of("Active uveitis") / cl_ref,
  "CRP 10 vs 0.6 mg/L CL/F ratio (Section 3.3: 'CL increased by 59%')", 1.59,  cl_of("CRP 10 mg/L") / cl_ref,
  "Allometry, 35 vs 70 kg (Table 2 exponent 0.75, i.e. 0.5^0.75)",    0.5946,  cl_of("35 kg, otherwise reference") / cl_ref
) |>
  dplyr::mutate(`% diff` = 100 * (Model - Paper) / Paper)

knitr::kable(claims, digits = 4,
             caption = "Deterministic covariate-model checks against the values printed in Nassar-Sheikh Rashid 2024.")

# These are algebraic identities, not simulated statistics, so a tight bound is
# the correct gate here (see known-vignette-failure-patterns.md pattern 11).
# The tolerance is set by the paper's own rounding: 0.269 and 1.59 are printed
# to 3 significant figures, and 0.5946 is 0.5^0.75 rounded to 4.
stopifnot(max(abs(claims$`% diff`)) < 0.2)
```

# Virtual cohort

Individual data are not public. Three virtual cohorts are built below, each
matching a different set of published demographics. `set.seed()` seeds R's RNG
only; rxode2's simulation RNG is partitioned per solver thread, so the exact
cohort differs between a 2-thread CI runner and a 16-thread workstation. Every
assertion downstream is therefore written to hold for any cohort this model can
produce.

```{r cohort-helpers}
set.seed(20240417)
rxode2::rxSetSeed(20240417)

TAU     <- 14                       # days between doses (every other week)
N_DOSE  <- 12                       # 154 days of dosing; steady state is reached well before the last interval
DOSE_T  <- TAU * (seq_len(N_DOSE) - 1)
SS_FROM <- DOSE_T[N_DOSE]           # 154 -- start of the final dosing interval
SS_TO   <- SS_FROM + TAU            # 168
WEEK10  <- 70                       # 10 weeks: the trough the paper reports in Section 3.4

# Coarse over the whole course (day 70 included for the 10-week trough), dense
# over the final interval so the NCA resolves Cmax and the AUC accurately.
OBS_T <- sort(unique(c(seq(0, SS_TO, by = 2), seq(SS_FROM, SS_TO, by = 0.25))))

# Approved European JIA dose band (Section 1 and reference [10]).
dose_for_wt <- function(wt) ifelse(wt < 30, 20, 40)

# Body weight from a lognormal matched to a reported median and IQR.
rlnorm_median_iqr <- function(n, med, q1, q3, lo, hi) {
  sdlog <- log(q3 / q1) / (2 * stats::qnorm(0.75))
  pmin(pmax(stats::rlnorm(n, meanlog = log(med), sdlog = sdlog), lo), hi)
}

# CRP is far too skewed for a lognormal (an IQR of 0.3-10.6 around a median of
# 4.0 implies sdlog ~ 2.6, which would put the 99th percentile near 2000 mg/L).
# Instead sample by inverse-CDF through the reported quartiles, with assumed
# min / max endpoints -- see "Assumptions and deviations".
rcrp <- function(n, q1, med, q3, lo, hi) {
  stats::approx(x = c(0, 0.25, 0.5, 0.75, 1), y = c(lo, q1, med, q3, hi),
                xout = stats::runif(n))$y
}

make_events <- function(subj) {
  dplyr::bind_rows(
    tidyr::crossing(id = subj$id, time = DOSE_T) |>
      dplyr::mutate(evid = 1L, cmt = "depot", rate = -2),
    tidyr::crossing(id = subj$id, time = OBS_T) |>
      dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", rate = 0)
  ) |>
    dplyr::left_join(subj, by = "id") |>
    dplyr::mutate(amt = ifelse(evid == 1L, dose_mg, amt)) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}
```

## Cohort A -- the paper's simulated JIA population (Section 3.4)

Section 3.4 describes the 500-patient virtual data set used for the paper's own
simulations: median body weight 45.3 kg (IQR 37.5-53.9), 13.2% with detectable
ADA, 77.4% on methotrexate, 15.4% with uveitis, median CRP 4.0 mg/L
(IQR 0.3-10.6). 200 subjects are used here (the library cap per arm); that is
ample for the median and mean statistics compared below.

```{r cohort-a}
N_A <- 200
subj_a <- tibble::tibble(
  id          = seq_len(N_A),
  WT          = rlnorm_median_iqr(N_A, 45.3, 37.5, 53.9, lo = 10, hi = 90),
  ADA_POS     = stats::rbinom(N_A, 1, 0.132),
  CONMED_MTX  = stats::rbinom(N_A, 1, 0.774),
  DIS_UVEITIS = stats::rbinom(N_A, 1, 0.154),
  CRP         = rcrp(N_A, q1 = 0.3, med = 4.0, q3 = 10.6, lo = 0.1, hi = 60)
) |>
  dplyr::mutate(dose_mg = dose_for_wt(WT), arm = "Simulated JIA cohort")

ev_a <- make_events(subj_a)
# No unique() here: anyDuplicated() applied to an already-deduplicated frame is
# always 0, i.e. a gate that cannot go red. The frame itself is checked.
stopifnot(!anyDuplicated(ev_a[, c("id", "time", "evid")]))
```

## Cohort B -- body-weight strata (Figure 2A)

Figure 2A stratifies the virtual population into five body-weight categories.
Each stratum below draws body weight uniformly within its band and the remaining
covariates from the Cohort A distributions, so the strata differ only in weight
(and therefore in the approved dose band).

```{r cohort-b}
N_B <- 100
bands <- tibble::tribble(
  ~band,      ~lo, ~hi,
  "<30 kg",    15,  30,
  "30-40 kg",  30,  40,
  "40-50 kg",  40,  50,
  "50-60 kg",  50,  60,
  ">60 kg",    60,  80
)

subj_b <- dplyr::bind_rows(lapply(seq_len(nrow(bands)), function(k) {
  tibble::tibble(
    id          = 1000L + (k - 1L) * N_B + seq_len(N_B),   # disjoint from cohorts A and C
    WT          = stats::runif(N_B, bands$lo[k], bands$hi[k]),
    ADA_POS     = stats::rbinom(N_B, 1, 0.132),
    CONMED_MTX  = stats::rbinom(N_B, 1, 0.774),
    DIS_UVEITIS = stats::rbinom(N_B, 1, 0.154),
    CRP         = rcrp(N_B, q1 = 0.3, med = 4.0, q3 = 10.6, lo = 0.1, hi = 60),
    arm         = bands$band[k]
  )
})) |>
  dplyr::mutate(dose_mg = dose_for_wt(WT),
                arm = factor(arm, levels = bands$band))

ev_b <- make_events(subj_b)
stopifnot(!anyDuplicated(ev_b[, c("id", "time", "evid")]))
```

## Cohort C -- the observed JIA cohort (Table 1)

Cohort C matches the *observed* patients rather than the paper's simulated ones:
median weight 49 kg (IQR 29.4-59.8), 18% ADA-positive, 78% on methotrexate, 16%
with uveitis, median CRP 0.55 mg/L (IQR 0.3-1.2). It exists so the model can be
checked against the observed median serum level of 12.0 mg/L (IQR 6.12-15.8),
which -- unlike the paper's simulated trough -- came from samples drawn at
arbitrary times within the dosing interval.

```{r cohort-c}
N_C <- 200
subj_c <- tibble::tibble(
  id          = 5000L + seq_len(N_C),
  WT          = rlnorm_median_iqr(N_C, 49, 29.4, 59.8, lo = 10, hi = 90),
  ADA_POS     = stats::rbinom(N_C, 1, 0.18),
  CONMED_MTX  = stats::rbinom(N_C, 1, 0.78),
  DIS_UVEITIS = stats::rbinom(N_C, 1, 0.16),
  CRP         = rcrp(N_C, q1 = 0.3, med = 0.55, q3 = 1.2, lo = 0.1, hi = 20)
) |>
  dplyr::mutate(dose_mg = dose_for_wt(WT), arm = "Observed-cohort demographics")

ev_c <- make_events(subj_c)
stopifnot(!anyDuplicated(ev_c[, c("id", "time", "evid")]))
```

# Simulation

```{r simulate}
solve_pop <- function(ev) {
  # omega is passed explicitly because the earlier zeroRe() solve of the same
  # compiled system can leave a zero omega in the solve options and silently
  # collapse every subject onto one typical patient.
  rxode2::rxSolve(
    ui, events = ev, omega = ui$omega,
    keep = c("arm", "WT", "ADA_POS", "CONMED_MTX", "DIS_UVEITIS", "CRP", "dose_mg"),
    returnType = "data.frame"
  )
}

sim_a <- solve_pop(ev_a)
sim_b <- solve_pop(ev_b)
sim_c <- solve_pop(ev_c)

# Guard the opposite failure to the one above: IIV must actually be sampled.
stopifnot(dplyr::n_distinct(round(sim_a$cl, 8)) > 1L)
# Cc is the individual prediction WITHOUT residual error. Everything below uses
# Cc deliberately; the additive residual is folded in explicitly where the
# paper's statistic includes it.
stopifnot(isTRUE(all.equal(sim_a$Cc, sim_a$ipredSim)))
stopifnot(all(sim_a$Cc >= 0), all(sim_b$Cc >= 0), all(sim_c$Cc >= 0))
```

# Replicate published figures

## Supplementary Figure 4A -- concentration-time profile to steady state

```{r figure-s4a}
sim_a |>
  dplyr::filter(time <= WEEK10) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    Q05 = stats::quantile(Cc, 0.05), Q50 = stats::median(Cc),
    Q95 = stats::quantile(Cc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 5, linetype = "dashed", colour = "grey40") +
  labs(x = "Time (days)", y = "Adalimumab concentration (mg/L)",
       title = "First five dosing intervals in the simulated JIA cohort",
       subtitle = "Median with 5th-95th percentile band; dashed line = the 5 mg/L adult RA target trough",
       caption = "Replicates Supplementary Figure 4A of Nassar-Sheikh Rashid 2024.")
```

## Figure 2A -- body-weight strata

```{r figure-2a}
ss_b <- sim_b |> dplyr::filter(time >= SS_FROM, time <= SS_TO)

ss_b |>
  dplyr::group_by(arm, t_int = time - SS_FROM) |>
  dplyr::summarise(Q50 = stats::median(Cc), .groups = "drop") |>
  ggplot(aes(t_int, Q50, colour = arm)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time within the dosing interval (days)",
       y = "Median adalimumab concentration (mg/L)",
       colour = "Body weight",
       title = "Figure 2A -- median steady-state profile by body-weight stratum",
       subtitle = "20 mg every other week below 30 kg, 40 mg every other week at or above 30 kg",
       caption = "Replicates Figure 2A of Nassar-Sheikh Rashid 2024.")

wt_median <- ss_b |>
  dplyr::group_by(arm) |>
  dplyr::summarise(`Median Cc over the interval (mg/L)` = stats::median(Cc),
                   .groups = "drop")
knitr::kable(wt_median, digits = 2,
             caption = "Median steady-state concentration by body-weight stratum.")

ratio_30_40_vs_60 <- wt_median[[2]][wt_median$arm == "30-40 kg"] /
  wt_median[[2]][wt_median$arm == ">60 kg"]
```

The 30-40 kg stratum reaches
`r sprintf("%.2f", ratio_30_40_vs_60)`-fold the concentrations of the above-60 kg
stratum. The paper describes this contrast as "nearly twice as high"
(Section 3.4); the model reproduces the direction and approximate magnitude of
the published claim.

```{r figure-2a-gate}
# A cohort-derived ratio, so the bound is a magnitude band rather than a value
# read off one run. Realised 1.64 on this render; a mis-transcribed allometric
# exponent or dose band moves this well outside [1.25, 2.60].
stopifnot(ratio_30_40_vs_60 > 1.25, ratio_30_40_vs_60 < 2.60)
# Section 3.4: patients nearer the 30 kg threshold have the highest exposure,
# so the top stratum by concentration must be one of the two lightest 40 mg
# bands rather than a heavy one.
stopifnot(as.character(wt_median$arm[which.max(wt_median[[2]])]) %in%
            c("30-40 kg", "40-50 kg"))
```

## Figure 2B-E -- covariate strata

```{r figure-2be}
ss_a <- sim_a |> dplyr::filter(time >= SS_FROM, time <= SS_TO)

strata <- dplyr::bind_rows(
  ss_a |> dplyr::transmute(t_int = time - SS_FROM, Cc,
                           panel = "B. Uveitis",
                           level = ifelse(DIS_UVEITIS == 1, "Uveitis", "No uveitis")),
  ss_a |> dplyr::transmute(t_int = time - SS_FROM, Cc,
                           panel = "C. Methotrexate",
                           level = ifelse(CONMED_MTX == 1, "On MTX", "No MTX")),
  ss_a |> dplyr::transmute(t_int = time - SS_FROM, Cc,
                           panel = "D. Anti-drug antibodies",
                           level = ifelse(ADA_POS == 1, "ADA-positive", "ADA-negative")),
  ss_a |> dplyr::transmute(t_int = time - SS_FROM, Cc,
                           panel = "E. C-reactive protein",
                           level = ifelse(CRP < 1, "CRP < 1 mg/L", "CRP >= 1 mg/L"))
)

strata |>
  dplyr::group_by(panel, level, t_int) |>
  dplyr::summarise(Q50 = stats::median(Cc), .groups = "drop") |>
  ggplot(aes(t_int, Q50, colour = level)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~panel, scales = "free_y") +
  labs(x = "Time within the dosing interval (days)",
       y = "Median adalimumab concentration (mg/L)", colour = NULL,
       title = "Figure 2B-E -- median steady-state profile by covariate status",
       caption = "Replicates Figure 2B-E of Nassar-Sheikh Rashid 2024.")
```

```{r figure-2be-gate}
# Section 3.4: "drug concentrations were highest in methotrexate users without
# detectable ADA, without uveitis during adalimumab treatment and with CRP
# levels below 1 mg/L". Each of the four contrasts below is driven by a
# covariate coefficient of at least 28% on clearance, so the ORDERING is a real
# structural claim rather than a near-zero effect whose sign could flip
# (contrast pattern 12 of known-vignette-failure-patterns.md).
panel_dir <- strata |>
  dplyr::group_by(panel, level) |>
  dplyr::summarise(med = stats::median(Cc), .groups = "drop_last") |>
  dplyr::summarise(
    favoured = level[which.max(med)],
    ratio    = max(med) / min(med),
    .groups  = "drop"
  )
knitr::kable(panel_dir, digits = 2,
             caption = "Higher-exposure group in each covariate panel, and the ratio between groups.")

stopifnot(identical(
  panel_dir$favoured[order(panel_dir$panel)],
  c("No uveitis", "On MTX", "ADA-negative", "CRP < 1 mg/L")
))
```

# PKNCA validation

Steady-state NCA is run over the final dosing interval (day
`r SS_FROM` to `r SS_TO`), by which point the cohort is at steady state: the
terminal half-life at the reference covariate set is about 13 days, and the
final interval starts after 11 doses.

```{r pknca}
nca_for <- function(sim, ev) {
  sim_nca <- sim |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::select(id, time, Cc, arm) |>
    dplyr::mutate(arm = as.character(arm))

  # Time-zero guarantee (pknca-recipes.md); pre-dose Cc = 0 is correct for a
  # subcutaneous model. Any existing time-zero row wins.
  sim_nca <- dplyr::bind_rows(
    sim_nca,
    sim_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
  ) |>
    dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
    dplyr::arrange(id, arm, time)

  dose_df <- ev |>
    dplyr::filter(evid == 1) |>
    dplyr::select(id, time, amt, arm) |>
    dplyr::mutate(arm = as.character(arm))

  conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id,
                               concu = "mg/L", timeu = "day")
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id,
                               doseu = "mg")

  # `cmin` rather than `ctrough`: PKNCA re-bases the time vector to the dose
  # before calling pk.calc.ctrough(), so `time %in% end` never matches and
  # ctrough comes back NA for every subject on a dose-anchored interval.
  #
  # On a dose-anchored steady-state interval the minimum falls at the START of
  # the interval, i.e. the pre-dose concentration immediately before the dose
  # that opens it. That is exactly the "trough level" the paper reports. It is
  # NOT the end-of-interval value: the ka distribution is wide (74.2% CV, fixed
  # from Kang) and its lower tail sits below kel, so a slow-absorbing subject
  # carries enough depot from earlier doses that central rises monotonically
  # across the interval. The chunk below asserts the identity rather than
  # assuming it.
  intervals <- data.frame(
    start = SS_FROM, end = SS_TO,
    cmax = TRUE, tmax = TRUE, cmin = TRUE, cav = TRUE, auclast = TRUE
  )
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}

nca_a <- nca_for(sim_a, ev_a)
nca_b <- nca_for(sim_b, ev_b)

res_a <- as.data.frame(nca_a$result)
res_b <- as.data.frame(nca_b$result)
stopifnot(nrow(res_a) > 0, nrow(res_b) > 0)   # a gate that cannot go red is worse than none

# Confirm `cmin` really is the pre-dose trough, so the comparison against the
# paper's trough below is like-for-like.
cmin_chk <- res_a |>
  dplyr::filter(PPTESTCD == "cmin") |>
  dplyr::transmute(id = as.integer(as.character(id)), cmin = PPORRES) |>
  dplyr::inner_join(
    sim_a |> dplyr::filter(time == SS_FROM) |> dplyr::select(id, cc_pre = Cc),
    by = "id"
  ) |>
  dplyr::inner_join(
    sim_a |> dplyr::filter(time == SS_TO) |> dplyr::select(id, cc_end = Cc),
    by = "id"
  )
stopifnot(nrow(cmin_chk) == N_A)

# Both sides are the same solved trajectory, so this is an exact identity and a
# tight bound is the correct gate: PKNCA's cmin must BE the pre-dose value.
stopifnot(max(abs(cmin_chk$cmin - cmin_chk$cc_pre)) < 1e-8)

# The pre-dose and end-of-interval troughs differ only by the accumulation still
# outstanding after 11 doses. Asserted on the centre and a robust quantile
# rather than the maximum: which subjects draw a slow ka -- and so are furthest
# from steady state -- is not reproducible across rxode2 builds or thread
# counts. Realised on one render: median 0.008%, 90th percentile 0.5%,
# maximum 4.1%.
ss_gap <- abs(cmin_chk$cc_pre - cmin_chk$cc_end) / cmin_chk$cc_end
stopifnot(stats::median(ss_gap) < 0.01,
          stats::quantile(ss_gap, 0.90) < 0.05)
```

## Internal identity: AUC over the interval equals dose / CL at steady state

At steady state the area under one dosing interval must equal `Dose / CL` for
every subject, whatever their absorption or distribution parameters. This check
is independent of the paper and validates that the ODE system, the `f(depot)`
anchor and the covariate model on clearance are wired together correctly.

```{r auc-identity}
cl_by_id <- sim_a |> dplyr::group_by(id) |>
  dplyr::summarise(cl = dplyr::first(cl), dose_mg = dplyr::first(dose_mg),
                   .groups = "drop")

auc_chk <- res_a |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(id, auc = PPORRES) |>
  dplyr::mutate(id = as.integer(as.character(id))) |>
  dplyr::inner_join(cl_by_id, by = "id") |>
  dplyr::mutate(pct_diff = 100 * (auc - dose_mg / cl) / (dose_mg / cl))
stopifnot(nrow(auc_chk) == N_A)

tibble::tibble(
  Statistic = c("Median % difference", "90th percentile of |% difference|",
                "Maximum |% difference|"),
  Value = c(stats::median(auc_chk$pct_diff),
            stats::quantile(abs(auc_chk$pct_diff), 0.9),
            max(abs(auc_chk$pct_diff)))
) |>
  knitr::kable(digits = 3, caption = "AUC over the final dosing interval versus Dose / CL.")

# Both sides use the SAME drawn parameters, so the residual difference is pure
# numerical error: trapezoidal integration on a 0.25-day grid, plus the last
# fraction of a percent of accumulation still outstanding after 11 doses. The
# central statistics are therefore held tightly; the upper tail is looser
# because the fixed 165% CV on Q/F produces occasional subjects whose
# distribution phase is still equilibrating. Realised on this render: median
# -0.015%, 90th percentile 1.24%, 95th percentile 2.62% (maximum 25.6%, driven
# by a single still-equilibrating subject). The bounds stop at the 95th
# percentile deliberately: over a 200-subject cohort the 99th percentile is the
# second-largest value, an extreme order statistic that is not reproducible
# across rxode2 builds or thread counts. Any of a wrong bioavailability, a
# wrong dose band or a mis-scaled clearance moves the MEDIAN by tens of percent
# and breaks the first bound immediately.
stopifnot(abs(stats::median(auc_chk$pct_diff)) < 2,
          stats::quantile(abs(auc_chk$pct_diff), 0.90) < 4,
          stats::quantile(abs(auc_chk$pct_diff), 0.95) < 10)
```

## Comparison against the published values

```{r nca-compare}
published <- tibble::tibble(arm = "Simulated JIA cohort", cmin = 7.4)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_a,
  reference = published,
  by        = "arm",
  units     = c(cmin = "mg/L"),
  tolerance_pct = 20
)

knitr::kable(cmp, caption = paste(
  "Simulated steady-state trough (PKNCA Cmin over the final dosing interval,",
  "which the chunk above confirms is the pre-dose trough) versus the",
  "value reported in Section 3.4 of Nassar-Sheikh Rashid 2024: 'the mean",
  "trough level in the simulated dataset was 7.4 +/- 5.5 mg/L'.",
  "* marks a difference above 20%."
))
```

```{r trough-detail}
trough_a <- sim_a |> dplyr::filter(time == WEEK10) |> dplyr::pull(Cc)
stopifnot(length(trough_a) == N_A)

# The paper's simulated data set is a NONMEM $SIMULATION, so its reported SD
# includes the additive residual error (SD 3.61 mg/L) on top of between-subject
# variability. Cc carries only the latter, so the comparable SD is the
# quadrature sum.
sd_ipred <- stats::sd(trough_a)
sd_total <- sqrt(sd_ipred^2 + 3.61^2)

tibble::tibble(
  Quantity = c("Mean 10-week trough (mg/L)",
               "SD of trough, between-subject only (mg/L)",
               "SD of trough, including the additive residual (mg/L)"),
  Paper    = c(7.4, NA_real_, 5.5),
  Model    = c(mean(trough_a), sd_ipred, sd_total)
) |>
  knitr::kable(digits = 2, caption = paste(
    "Ten-week (five-interval) trough in the simulated JIA cohort, against",
    "Section 3.4 of Nassar-Sheikh Rashid 2024."
  ))

# A cohort mean over 200 subjects with a right-skewed clearance distribution.
# The band is wide enough to absorb the draw-to-draw spread but narrow enough
# that a mis-transcribed clearance, dose band or dosing interval -- each of
# which moves the trough by tens of percent -- still breaks it.
stopifnot(abs(mean(trough_a) / 7.4 - 1) < 0.35)
# The paper's SD is an envelope claim, not a point estimate; require only that
# the model's total variability is of the same order.
stopifnot(sd_total > 0.5 * 5.5, sd_total < 2.5 * 5.5)
```

## Observed concentrations (Table 1)

The observed cohort's samples were drawn at arbitrary times within the dosing
interval, so the comparable model quantity is a single randomly-timed
steady-state sample per subject, not a trough.

```{r observed-compare}
set.seed(99)
# One randomly-timed sample per subject, drawn once in base R and joined on, so
# the sampling times are fixed by set.seed() rather than redrawn per group.
rand_t <- subj_c |>
  dplyr::transmute(id, t_target = SS_FROM + stats::runif(dplyr::n(), 0, TAU))

sample_c <- sim_c |>
  dplyr::filter(time >= SS_FROM, time <= SS_TO) |>
  dplyr::inner_join(rand_t, by = "id") |>
  dplyr::group_by(id) |>
  dplyr::slice(which.min(abs(time - t_target))) |>
  dplyr::ungroup()
stopifnot(nrow(sample_c) == N_C)

obs_cmp <- tibble::tibble(
  Statistic = c("Median (mg/L)", "25th percentile (mg/L)", "75th percentile (mg/L)"),
  `Observed (Table 1)` = c(12.0, 6.12, 15.8),
  Model = c(stats::median(sample_c$Cc),
            stats::quantile(sample_c$Cc, 0.25),
            stats::quantile(sample_c$Cc, 0.75))
)
knitr::kable(obs_cmp, digits = 2, caption = paste(
  "Randomly-timed steady-state samples from a virtual cohort matching the",
  "observed Table 1 demographics, against the observed adalimumab levels."
))

# The model is being asked to predict a real cohort's concentration
# distribution from covariates alone, so this is a structural sanity band, not
# a precision claim. It still fails loudly on a mis-scaled clearance or volume.
stopifnot(abs(stats::median(sample_c$Cc) / 12.0 - 1) < 0.40)
```

The lower quartile lands almost exactly on the observed 6.12 mg/L, but the
model's median and upper quartile sit roughly 15-25% below the observed values.
This is a real, reproducible under-prediction rather than a transcription
problem -- the deterministic covariate checks above reproduce every printed
number to better than 0.2% -- and there are three plausible contributors, all
of which the source itself raises:

1. **The imputed ADA effect biases clearance upward.** Section 4 states that the
   drug-sensitive assay used here under-detects ADA-positive patients, so
   "correcting too little for the presence of ADA could have resulted in an
   overestimation of clearance". An over-estimated `CL/F` under-predicts
   concentrations, which is the direction seen here.
2. **Sampling times are ambiguous in the source.** Section 2.1 describes the
   samples as trough concentrations, while Section 3.1 says they were "collected
   at random time points in the treatment interval". Random-time sampling is
   assumed above; assuming troughs instead would move the model's median further
   down, not up, so the discrepancy is not explained by this choice alone.
3. **Ascertainment.** Levels were measured "mainly in case of clinical
   ineffectiveness" (Section 4), which biases the observed sample in a
   direction the model has no way to represent.

The gate above is therefore a structural band rather than a precision claim, and
the deviation is recorded rather than tuned away.

# Assumptions and deviations

- **Known deviation: the observed median is under-predicted by roughly 20%.**
  The model reproduces the paper's own simulated trough (7.4 mg/L) and every
  printed covariate multiplier essentially exactly, but a virtual cohort matching
  the *observed* Table 1 demographics predicts a median randomly-timed level
  below the observed 12.0 mg/L. See the "Observed concentrations" section above
  for the three contributing mechanisms the source itself raises. This is
  recorded, not tuned away, and the corresponding check is a structural band
  rather than a precision gate.
- **PKNCA `ctrough` is unusable on a dose-anchored interval.** PKNCA re-bases the
  time vector to the dose before calling `pk.calc.ctrough()`, so `time %in% end`
  never matches and every subject returns `NA`. `cmin` is used instead, with an
  explicit assertion identifying which trough it is: on a steady-state interval
  anchored at a dose, the minimum falls at the *start* of the interval -- the
  pre-dose concentration -- for every subject, and that is the quantity the
  paper's "trough level" refers to. It is deliberately not compared against the
  end-of-interval value: the fixed 74.2% CV on `ka` puts the lower tail of the
  absorption-rate distribution below `kel`, so slow-absorbing subjects carry
  enough depot from earlier doses that concentration rises monotonically across
  the interval.
- **Omega off-diagonals are not from this paper.** Table 2 prints only the
  diagonal %CV values for the four fixed inter-individual variances. Section 3.3
  says the authors kept "the full omega variance-covariance matrix for the other
  parameters as described previously", so the six covariances in the fixed
  `BLOCK(4)` were read from the corresponding sub-block of Kang 2020 Table 4 via
  the registered upstream extraction
  `modellib("Kang_2020_adalimumab_phase3_extension")`. Dropping them would
  contradict the paper's own description of the model; the resulting 4x4 is
  positive definite (smallest eigenvalue 0.064). The `etalcl` variance sits
  outside that block because it is the one variance this paper re-estimated.
- **Bioavailability is a structural anchor.** No intravenous arm was studied and
  every disposition parameter in Table 2 is apparent (CL/F, V1/F, Q/F, V2/F), so
  F is unidentifiable and is carried at 1.
- **The ADA effect is imputed, not fitted.** `theta_ADA = 2.08` was fixed rather
  than estimated because the JIA data could not identify it. The paper derives
  it from Kang 2020 model 3, whose reference group is ADA-*positive*, by
  inverting the 0.654 ADA-negative factor and then doubling the resulting effect
  to absorb Kang's separate ADA-titre term. Note it is the fractional *increase*
  that is doubled, not the ratio: inverting 0.65 gives a 53% higher clearance,
  and 2 x 53% = 106%, i.e. the factor of about 2.08 that Table 2 reports and the
  "108% increase" Section 3.3 quotes. Doubling the ratio itself would give 3.06,
  which is not the reported value. The authors flag this in Section 4 as a limitation:
  assay-to-assay differences in ADA detection may make the imputed effect wrong,
  which would in turn bias the estimated clearance.
- **CRP reference value.** The model uses the 0.6 mg/L reference printed in the
  Table 2 row label and in the Section 3.3 worked example, not the 0.55 mg/L
  median reported in Table 1; the paper evidently rounded. The 0.165 exponent
  is only supported over the narrow CRP range this cohort spans (IQR
  0.3-1.2 mg/L observed) and up to the 10 mg/L used in the paper's own example.
- **CRP sampling distribution in the vignette.** The simulated cohort's CRP is
  drawn by inverse-CDF interpolation through the reported quartiles rather than
  from a lognormal: an IQR of 0.3-10.6 mg/L around a median of 4.0 implies
  `sdlog` of about 2.6, which would generate physiologically absurd values. The
  minimum (0.1 mg/L) and maximum (60 mg/L for the simulated cohort, 20 mg/L for
  the observed cohort) endpoints are assumed, not published.
- **Body-weight distributions are assumed lognormal**, matched to the published
  median and interquartile range and clamped to 10-90 kg. Only the median and
  IQR were reported.
- **Weight-band boundaries in Figure 2A.** The paper says the virtual population
  was split into five body-weight categories and names two of them ("30-40 kg",
  "over 60 kg"). The intermediate cut-points used here (<30, 30-40, 40-50,
  50-60, >60) are inferred from those two plus the 30 kg dose threshold.
- **Race and ethnicity are not reported** in the source and are absent from the
  model's `population` metadata and from the virtual cohorts.
- **Covariates are treated as time-fixed in the simulations.** In the source,
  CRP and ADA status are time-varying with last-observation-carried-forward
  imputation (Section 2.5); a single steady-state value per subject is used here
  because the validation targets are steady-state statistics.
- **Screened but unretained covariates** (sex, body surface area, and any
  concomitant immunosuppressive medication) are recorded in the model file's
  `covariatesDataExcluded` list rather than `covariateData`, since they are not
  referenced in `model()`.
- **The sensitivity analysis is not extracted.** Supplementary Table 3 reports
  the final model refitted without CRP; per the library's replicate-the-authors
  policy, robustness checks that the authors did not report as the final model
  are excluded.
- **The five literature comparator models are not re-extracted here.** They were
  evaluated, not developed, by this paper. Three of them are already in
  nlmixr2lib as `Kang_2020_adalimumab_phase1`,
  `Kang_2020_adalimumab_phase3_base` and
  `Kang_2020_adalimumab_phase3_extension`.
