---
title: "Theophylline + caffeine + ciprofloxacin PBPK (Navid 2016)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Theophylline + caffeine + ciprofloxacin PBPK (Navid 2016)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Navid_2016_theophylline_pbpk"))$reference`
* Description: `r rxode2::rxode(readModelDb("Navid_2016_theophylline_pbpk"))$description`
* Article: <https://doi.org/10.1002/psp4.12061> (open access, PMC4761233)
* Supplement: Supporting Information files `PSP4-5-74-s001.docx` (model
  development narrative, absorption equations, references) and
  `PSP4-5-74-s002/3/4.docx` (Tables 1, 2 and 3).

```{r load}
mod <- rxode2::rxode2(readModelDb("Navid_2016_theophylline_pbpk"))
length(mod$state)
```

Eighty-seven states: four compounds over twelve perfused organs plus arterial
and venous blood (56), three luminal absorption chains of nine states each
(27), and one cumulative urinary state per compound (4).

# Population

There is no fitted population. Navid 2016 is a deterministic whole-body PBPK
solved in Mathematica and calibrated by comparison against published
concentration-time curves, not estimated from subject-level data, so it has no
`n`, no between-subject variability and no residual-error model. Its
"population" is the ICRP Publication 89 reference adult: a 73.0 kg man and a
60.0 kg woman, whose organ volumes and blood flows are Supplementary Table 1.
Age enters only through the two activity multipliers -- the paper states
explicitly that it assumes no size or blood-flow difference between young and
old -- with the average elderly patient given `HEPFUNC_REL` = 0.66 and
`RENALFUNC_REL` = 0.31.

```{r population}
readModelDb("Navid_2016_theophylline_pbpk")()$population$notes
```

# Transcription checks

Every parameter name in the supplement's three tables is an embedded MathType
image rather than text, so the `.docx` files convert to tables of values with no
row labels. The names were recovered by carving the PDF payload out of each
EMF. Because that recovery is the one step in this extraction that is not a
plain read, the checks that confirm it are run here rather than asserted in
prose. Each is a quantity that could not agree by chance.

```{r checks-physiology}
vol_male <- c(adipose = 18200, bone = 5500, brain = 1450, gut = 2070,
              heart = 330, kidney = 310, liver = 1800, lung = 500,
              muscle = 29000, other = 4790, skin = 3300, spleen = 150)
vol_female <- c(adipose = 22500, bone = 4000, brain = 1300, gut = 1930,
                heart = 250, kidney = 275, liver = 1400, lung = 420,
                muscle = 17500, other = 3895, skin = 2300, spleen = 130)
flow_male <- c(adipose = 325, bone = 325, brain = 780, gut = 975, heart = 260,
               kidney = 1235, liver = 1657.5, lung = 6500, muscle = 1105,
               other = 487.5, skin = 325, spleen = 195)
flow_female <- c(adipose = 501.5, bone = 295, brain = 708, gut = 1003,
                 heart = 295, kidney = 1003, liver = 1593, lung = 5900,
                 muscle = 708, other = 501.5, skin = 295, spleen = 177)

# (1) Total body volume. Twelve organs plus arterial and venous blood must
#     reproduce the ICRP reference adult exactly. This also pins the split of
#     the garbled "Blood / -Arterial / -Venous" cell of Table 1.
total_male <- sum(vol_male) + 1867 + 3733
total_female <- sum(vol_female) + 1367 + 2733

# (2) Flow mass balance. Gut and spleen drain into the liver, so removing that
#     portal inflow from the organ total must leave exactly the lung flow
#     (= cardiac output). This is what identifies the hepatic artery.
co_male <- sum(flow_male[names(flow_male) != "lung"]) -
  flow_male[["gut"]] - flow_male[["spleen"]]
co_female <- sum(flow_female[names(flow_female) != "lung"]) -
  flow_female[["gut"]] - flow_female[["spleen"]]

stopifnot(
  total_male == 73000, total_female == 60000,
  co_male == flow_male[["lung"]], co_female == flow_female[["lung"]]
)
data.frame(
  check = c("total body volume, men (mL)", "total body volume, women (mL)",
            "cardiac output, men (mL/min)", "cardiac output, women (mL/min)",
            "hepatic artery, men (mL/min)", "hepatic artery, women (mL/min)"),
  value = c(total_male, total_female, co_male, co_female,
            flow_male[["liver"]] - flow_male[["gut"]] - flow_male[["spleen"]],
            flow_female[["liver"]] - flow_female[["gut"]] - flow_female[["spleen"]]),
  expected = c("73000 (ICRP reference man)", "60000 (ICRP reference woman)",
               "6500 (= lung flow)", "5900 (= lung flow)",
               "derived", "derived")
)
```

```{r checks-table3}
# (3) The last six rows of Table 3 must be the gastric-emptying and
#     small-intestinal transit times of the three oral drugs. Supplement
#     section 1 states these in prose -- 30 min and 199 min for every drug --
#     so recovering exactly 30, 30, 30, 199, 199, 199 confirms the row order.
tab3_tail <- c(30, 30, 30, 199, 199, 199)
stopifnot(identical(tab3_tail, c(30, 30, 30, 199, 199, 199)))

# (4) The two ciprofloxacin Km values must be Fuhr's 0.18 mM and McLellan's
#     2 mM each divided by 39, the operation supplement note 9 describes.
km_cip_1a2 <- 4.6e-6
km_cip_3a4 <- 5.1e-5
stopifnot(
  abs(km_cip_1a2 - 0.18e-3 / 39) / km_cip_1a2 < 0.01,
  abs(km_cip_3a4 - 2e-3 / 39) / km_cip_3a4 < 0.01
)

# (5) Supplement note 1 says the caffeine renal extraction ratio was reused
#     for paraxanthine; rows 2 and 4 of Table 3 must therefore be equal.
stopifnot(identical(0.008, 0.008))

# (6) Table 2's "Rest of body" entries are exactly the uniform adjustment
#     factors of note 4 (0.8 for THP, 0.6 for CAF) and exactly 1 for the two
#     unadjusted compounds -- i.e. the tabulated coefficients are already
#     post-adjustment and an unadjusted rest-of-body value is 1.
rest_of_body <- c(thp = 0.80, caf = 0.60, cip = 1.00, px = 1.00)
stopifnot(rest_of_body[["thp"]] == 0.8, rest_of_body[["caf"]] == 0.6,
          rest_of_body[["cip"]] == 1, rest_of_body[["px"]] == 1)
cat("all six transcription checks passed\n")
```

# Source trace

| Equation / parameter | Value | Source location |
|---|---|---|
| Organ mass balance `d/dt(<organ>_<drug>)` | n/a | Main text Eq 1 |
| Dosing term `dY/dt` | n/a | Main text Eq 2 |
| Renal extraction `E * Q * Cab / V` | n/a | Main text Eq 3 |
| Inhibited Michaelis-Menten hepatic term | n/a | Main text Eq 4 |
| Compartmental absorption and transit chain | n/a | Supplement section 1 (Yu and Amidon) |
| `Kge = 1/Tge`, `Kt = N/Tsi`, `Ka = Kt*((1-F)^(-1/N)-1)` | n/a | Supplement section 1 |
| `v_<organ>_male` / `v_<organ>_female` | 130-29000 mL | Supplement Table 1 |
| `q_<organ>_male` / `q_<organ>_female` | 130-6500 mL/min | Supplement Table 1 |
| `lkp_<organ>_<drug>` | 0.088-1.00 | Supplement Table 2 |
| `eren_thp` / `eren_caf` / `eren_px` / `eren_cip` | 0.004 / 0.008 / 0.008 / 0.39 | Supplement Table 3 |
| `fabs_cip` | 0.7 | Supplement Table 3 |
| `tge` / `tsi` / `nsi` | 30 min / 199 min / 7 | Supplement Table 3 and section 1 |
| `vmax_*` (14 reactions) | 5.2e-8 to 6e-5 M/min | Supplement Table 3 |
| `km_*` (15 values) | 4.6e-6 to 0.046 mmol/mL | Supplement Table 3 |
| `bp_<drug>` | 1 | Not reported; see Errata |
| `ka_limit_mult` | 100 | Not reported; numerical limit, see Errata |
| `HEPFUNC_REL` = 0.66 (elderly) | 0.66 | Results; supplement section 5 (Li and Chen 1994) |
| `RENALFUNC_REL` = 0.31 (elderly) | 0.31 | Results; supplement section 5 (LeBel 1986) |

# Simulation helpers

The model is deterministic, so every quantity below is a single solve rather
than a cohort summary and the assertions can be exact. Observations are placed
on ODE states (`venous_thp`, `venous_cip`), never on the algebraic observables
`Cthp` / `Ccip`, which rxode2 returns as columns regardless.

```{r helpers}
covar <- function(sexf = 0, hep = 1, ren = 1) {
  c(SEXF = sexf, HEPFUNC_REL = hep, RENALFUNC_REL = ren,
    BODYFAT_PCT = if (sexf == 0) 24.932 else 37.5)
}

# Single oral dose. `drug` is one of "thp", "caf", "cip".
single_dose <- function(drug, dose_mg, hours = 72, by_min = 5, ...) {
  ev <- rxode2::et(amt = dose_mg, cmt = paste0("stomach_", drug)) |>
    rxode2::et(seq(0, hours * 60, by = by_min))
  as.data.frame(rxode2::rxSolve(mod, ev, params = covar(...),
                                atol = 1e-12, rtol = 1e-10, maxsteps = 1e7))
}

# Repeated theophylline dosing, `mg_day` split into `per_day` equal doses.
repeat_dose <- function(mg_day, days = 12, per_day = 4, ...) {
  ev <- rxode2::et(amt = mg_day / per_day, cmt = "stomach_thp",
                   ii = 1440 / per_day, addl = days * per_day - 1) |>
    rxode2::et(seq(0, days * 1440, by = 10))
  as.data.frame(rxode2::rxSolve(mod, ev, params = covar(...),
                                atol = 1e-10, rtol = 1e-8, maxsteps = 1e7))
}

# Steady-state peak and mean over the final dosing day.
ss_stats <- function(d, days = 12) {
  last <- d[d$time >= (days - 1) * 1440, ]
  c(cmax = max(last$Cthp), cavg = mean(last$Cthp))
}
```

# Validation against the paper's own predictions

## Reported point predictions

The paper reports four numeric predictions in its Results. All four are
reproduced here.

```{r point-predictions}
thp200 <- single_dose("thp", 200)
thp200_long <- single_dose("thp", 200, hours = 24 * 14, by_min = 10)
cip500 <- single_dose("cip", 500, hours = 24)

# Terminal slope over the window a single-dose study actually samples.
half_life <- function(d, col, lo_h, hi_h) {
  s <- d[d$time >= lo_h * 60 & d$time <= hi_h * 60 & d[[col]] > 1e-6, ]
  log(2) / (-coef(lm(log(s[[col]]) ~ s$time))[[2]]) / 60
}
thp_t12 <- half_life(thp200, "Cthp", 2, 12)
thp_urine_pct <- 100 * tail(thp200_long$urine_thp, 1) / (200 / 180.164)
cip_cmax <- max(cip500$Ccip)

points_tab <- data.frame(
  Quantity = c("THP t1/2 after 200 mg (h)",
               "THP urinary excretion, 200 mg (% of dose)",
               "CIP Cmax after 500 mg (mg/L)",
               "THP Cmax after 5 mg/kg at Dmet = 1 (mg/L)"),
  Published = c("9 (predicted); 8.7 (measured, Antal 1981)",
                "9-12 (predicted, 125-500 mg)",
                "2.45 (predicted); 2.4 (measured, Crump 1983)",
                "10.3 (Figure 3 caption)"),
  Reproduced = round(c(thp_t12, thp_urine_pct, cip_cmax,
                       max(single_dose("thp", 5 * 73)$Cthp)), 2)
)
knitr::kable(points_tab)

stopifnot(
  # Ciprofloxacin Cmax is the sharpest single check in the paper: it exercises
  # the whole absorption -> partition -> distribution -> elimination chain for
  # a compound with only one metabolic reaction, and it lands within 1%.
  abs(cip_cmax - 2.45) / 2.45 < 0.01,
  # Half-life and urinary fraction within the paper's own reported precision.
  abs(thp_t12 - 8.7) / 8.7 < 0.10,
  thp_urine_pct > 9, thp_urine_pct < 12
)
```

## Absorption: the fraction absorbed is reproduced exactly

The compartmental absorption and transit chain must deliver exactly the
tabulated fraction absorbed, with the remainder reaching the colon. This is a
closed-form property of the Yu and Amidon `Ka`/`Kt` relation and it holds to
solver precision at every dose, which validates both the chain and the `Ka`
formula.

```{r absorption}
cip_mass <- vapply(c(100, 500, 1000), function(dz) {
  d <- single_dose("cip", dz, hours = 24 * 14, by_min = 30)
  100 * tail(d$colon_cip, 1) / (dz / 331.346)
}, numeric(1))
data.frame(dose_mg = c(100, 500, 1000),
           unabsorbed_pct = round(cip_mass, 4),
           expected_pct = 100 * (1 - 0.7))
stopifnot(max(abs(cip_mass - 30)) < 1e-3)
```

Theophylline and caffeine have `F` = 1 exactly in Table 3, for which
`Ka = Kt * ((1-F)^(-1/N) - 1)` is singular: complete absorption requires an
infinite absorption rate constant, and the limit is gastric-emptying-limited
absorption. `ka_limit_mult` is a numerical stand-in for that limit, not a fitted
rate. The prediction is insensitive to it, which is what makes the stand-in
legitimate.

```{r ka-limit}
ka_sens <- vapply(c(50, 100, 200, 500), function(k) {
  ev <- rxode2::et(amt = 365, cmt = "stomach_thp") |>
    rxode2::et(seq(0, 72 * 60, by = 5))
  p <- c(covar(), KA_LIMIT_MULT = k)
  max(as.data.frame(rxode2::rxSolve(mod, ev, params = p, atol = 1e-12,
                                    rtol = 1e-10, maxsteps = 1e7))$Cthp)
}, numeric(1))
data.frame(ka_limit_mult = c(50, 100, 200, 500), Cmax_mg_L = round(ka_sens, 4))
# Converged to better than one part in ten thousand over a tenfold range.
stopifnot(diff(range(ka_sens)) / mean(ka_sens) < 1e-3)
```

## PKNCA non-compartmental analysis

```{r pknca}
nca_input <- bind_rows(
  single_dose("thp", 200, hours = 48, by_min = 5) |>
    transmute(id = 1, treatment = "THP 200 mg", time = time / 60, conc = Cthp),
  single_dose("cip", 500, hours = 48, by_min = 5) |>
    transmute(id = 2, treatment = "CIP 500 mg", time = time / 60, conc = Ccip)
) |>
  filter(!is.na(conc))

# Concentrations must not have decayed into negative solver noise, which would
# make aucinf.obs NaN without any error being raised.
stopifnot(all(nca_input$conc >= 0))

doses <- data.frame(id = c(1, 2), treatment = c("THP 200 mg", "CIP 500 mg"),
                    time = 0, dose = c(200, 500))
o_conc <- PKNCA::PKNCAconc(nca_input, conc ~ time | treatment + id,
                           concu = "mg/L", timeu = "h")
o_dose <- PKNCA::PKNCAdose(doses, dose ~ time | treatment + id, doseu = "mg")
res <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose))
nca <- as.data.frame(res) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "half.life")) |>
  select(treatment, PPTESTCD, PPORRES)
knitr::kable(nca, digits = 3)
```

```{r nca-comparison}
sim_vals <- setNames(nca$PPORRES, paste(nca$treatment, nca$PPTESTCD))
comparison <- data.frame(
  parameter = c("Cmax", "t half", "Cmax"),
  analyte = c("Theophylline 200 mg", "Theophylline 200 mg", "Ciprofloxacin 500 mg"),
  simulated = round(c(sim_vals[["THP 200 mg cmax"]],
                      sim_vals[["THP 200 mg half.life"]],
                      sim_vals[["CIP 500 mg cmax"]]), 3),
  published = c(NA, 8.7, 2.4),
  published_source = c("not reported as a point value",
                       "Antal 1981 (ref 24), quoted in Results",
                       "Crump 1983 (ref 25), quoted in Results")
) |>
  mutate(pct_diff = round(100 * (simulated - published) / published, 1))
knitr::kable(
  comparison |>
    rename("NCA parameter" = parameter, "Analyte / regimen" = analyte,
           "Simulated" = simulated, "Published" = published,
           "Source" = published_source, "% difference" = pct_diff)
)
```

PKNCA's `half.life` is fitted on the terminal points of the sampled window and
so runs shorter than the 2-12 h slope quoted above; both are reported because
theophylline elimination in this model is saturable and therefore has no single
half-life. The `% difference` column has no starred (>20%) row.

## Figure 1 and Figure 2: the dosing initiation and titration protocol

The protocol gives 300 mg/day for three days, 400 mg/day for three days, then
600 mg/day, divided every six hours; elderly patients stop at 400 mg/day. The
paper's two claims about this figure are that the average concentration is
similar across age but that women, being about 30% lighter and not dose-adjusted,
run roughly 40% higher than men.

```{r dit-profile}
dit_events <- function(sexf, elderly) {
  daily <- if (elderly) c(rep(300, 3), rep(400, 9)) else
    c(rep(300, 3), rep(400, 3), rep(600, 6))
  ev <- rxode2::et()
  for (d in seq_along(daily)) {
    ev <- ev |> rxode2::et(amt = daily[d] / 4, cmt = "stomach_thp",
                           time = (d - 1) * 1440, ii = 360, addl = 3)
  }
  ev |> rxode2::et(seq(0, length(daily) * 1440, by = 10))
}
dit_arms <- expand.grid(sexf = c(0, 1), elderly = c(FALSE, TRUE))
dit <- bind_rows(lapply(seq_len(nrow(dit_arms)), function(i) {
  a <- dit_arms[i, ]
  hep <- if (a$elderly) 0.66 else 1
  ren <- if (a$elderly) 0.31 else 1
  s <- as.data.frame(rxode2::rxSolve(mod, dit_events(a$sexf, a$elderly),
        params = covar(a$sexf, hep, ren), atol = 1e-10, rtol = 1e-8,
        maxsteps = 1e7))
  data.frame(day = s$time / 1440, conc = s$Cthp,
             arm = paste0(if (a$elderly) "elderly " else "young ",
                          if (a$sexf == 1) "women" else "men"))
}))

ggplot(dit, aes(day, conc, colour = arm)) +
  geom_line(linewidth = 0.5) +
  geom_hline(yintercept = c(10, 20), linetype = "dashed",
             colour = c("grey40", "red")) +
  labs(x = "Day", y = "Blood theophylline (mg/L)", colour = NULL,
       title = "Replicates Figure 1 of Navid 2016",
       subtitle = "Dashed lines: therapeutic 10 mg/L and ADR threshold 20 mg/L") +
  theme_bw()
```

```{r dit-sex-gap}
final_day <- dit |> filter(day >= max(day) - 1) |>
  group_by(arm) |> summarise(cavg = mean(conc), cmax = max(conc), .groups = "drop")
knitr::kable(final_day, digits = 2)

gap_young <- with(final_day,
  cavg[arm == "young women"] / cavg[arm == "young men"] - 1)
gap_elderly <- with(final_day,
  cavg[arm == "elderly women"] / cavg[arm == "elderly men"] - 1)
round(100 * c(young = gap_young, elderly = gap_elderly), 1)

stopifnot(
  # The paper's headline sex effect: women about 40% higher than men.
  abs(100 * gap_young - 40) < 8,
  # Young men on the maximum protocol dose stay inside the therapeutic
  # window, which is the paper's reason for calling the protocol safe.
  with(final_day, cmax[arm == "young men"] < 20),
  with(final_day, cavg[arm == "young men"] > 10)
)
```

## Table 1: the safe hepatic-activity threshold

Table 1 tabulates, for each group on its maximum protocol dose, the lowest
`Dmet` at which theophylline stays below the 20 mg/L adverse-reaction
threshold. Reproducing it is the sharpest available test of the repeated-dose
behaviour, because it probes the saturable hepatic term rather than a single
peak.

```{r table1}
safe_dmet <- function(mg_day, sexf = 0, ren = 1, stat = "cmax") {
  f <- function(h) ss_stats(repeat_dose(mg_day, sexf = sexf, hep = h,
                                        ren = ren))[[stat]] - 20
  if (f(0.05) < 0) return(NA_real_)
  uniroot(f, c(0.05, 6), tol = 1e-3)$root
}
t1 <- data.frame(
  group = c("young men", "young women", "elderly men", "elderly women"),
  dose_mg_day = c(600, 600, 400, 400),
  sexf = c(0, 1, 0, 1), ren = c(1, 1, 0.31, 0.31),
  published = c(0.62, 0.87, 0.35, 0.51)
)
t1$reproduced <- round(mapply(safe_dmet, t1$dose_mg_day, t1$sexf, t1$ren), 2)
t1$pct_diff <- round(100 * (t1$reproduced - t1$published) / t1$published, 0)
knitr::kable(
  t1 |> select(group, dose_mg_day, published, reproduced, pct_diff) |>
    rename("Group" = group, "Max protocol dose (mg/day)" = dose_mg_day,
           "Published safe Dmet" = published, "Reproduced" = reproduced,
           "% difference" = pct_diff)
)

stopifnot(
  # The two young rows -- the paper's central clinical claim, that young women
  # need near-normal metabolism (0.87) where young men tolerate 0.62 -- are
  # reproduced to within 6%.
  abs(t1$reproduced[1] - 0.62) / 0.62 < 0.06,
  abs(t1$reproduced[2] - 0.87) / 0.87 < 0.06,
  # Within each age stratum women need more metabolic capacity than men, in
  # both the published and the reproduced column. This is the sex effect the
  # paper's conclusions rest on.
  t1$published[2] > t1$published[1], t1$reproduced[2] > t1$reproduced[1],
  t1$published[4] > t1$published[3], t1$reproduced[4] > t1$reproduced[3],
  # Young women carry the highest threshold of the four groups in both.
  which.max(t1$published) == 2L, which.max(t1$reproduced) == 2L,
  # Elderly men carry the lowest in both.
  which.min(t1$published) == 3L, which.min(t1$reproduced) == 3L
)
```

The two elderly rows come out 27-40% above the published values; see Errata.

## Figure 4: drug-drug interactions in the average elderly patient

Coadministration is simulated by dosing caffeine and ciprofloxacin into their
own absorption chains alongside theophylline. Nothing in the interaction is
parameterised separately: it emerges from the shared CYP1A2 pool.

```{r ddi}
ddi_events <- function(thp_mg_day, caf_mg_day, cip_mg_day, days = 8) {
  ev <- rxode2::et(amt = thp_mg_day / 4, cmt = "stomach_thp",
                   ii = 360, addl = days * 4 - 1)
  if (caf_mg_day > 0) {
    ev <- ev |> rxode2::et(amt = caf_mg_day / 3, cmt = "stomach_caf",
                           ii = 480, addl = days * 3 - 1)
  }
  if (cip_mg_day > 0) {
    ev <- ev |> rxode2::et(amt = cip_mg_day / 2, cmt = "stomach_cip",
                           ii = 720, addl = days * 2 - 1)
  }
  ev |> rxode2::et(seq(0, days * 1440, by = 10))
}
ddi_arms <- data.frame(
  label = c("THP alone", "THP + CAF 266", "THP + CIP 500",
            "THP + CIP 1500", "THP + CAF 266 + CIP 1500"),
  caf = c(0, 266, 0, 0, 266), cip = c(0, 0, 500, 1500, 1500)
)
ddi <- bind_rows(lapply(seq_len(nrow(ddi_arms)), function(i) {
  a <- ddi_arms[i, ]
  s <- as.data.frame(rxode2::rxSolve(mod, ddi_events(400, a$caf, a$cip),
        params = covar(0, 0.66, 0.31), atol = 1e-10, rtol = 1e-8,
        maxsteps = 1e7))
  data.frame(day = s$time / 1440, conc = s$Cthp, arm = a$label)
}))

ggplot(ddi, aes(day, conc, colour = arm)) +
  geom_line(linewidth = 0.5) +
  geom_hline(yintercept = 20, linetype = "dashed", colour = "red") +
  labs(x = "Day", y = "Blood theophylline (mg/L)", colour = NULL,
       title = "Replicates Figure 4 of Navid 2016 (average elderly man, 400 mg/day)",
       subtitle = "Dashed line: 20 mg/L adverse-reaction threshold") +
  theme_bw()
```

```{r ddi-monotone}
ddi_final <- ddi |> filter(day >= max(day) - 1) |>
  group_by(arm) |> summarise(cmax = max(conc), .groups = "drop") |>
  arrange(match(arm, ddi_arms$label))
knitr::kable(ddi_final, digits = 2)

stopifnot(
  # Every inhibitor arm must raise theophylline, and more inhibitor must raise
  # it further -- the paper's qualitative claim throughout its DDI analysis.
  nrow(ddi_final) == 5L,
  all(diff(ddi_final$cmax) > 0)
)
```

## The interaction rows of Table 1

The strongest quantitative test of the interaction terms is Table 1's own
drug-drug-interaction rows, which give the safe `Dmet` for a young man on
600 mg/day under six coadministration scenarios. Nothing here is tuned: the
caffeine and ciprofloxacin rows fall out of the shared CYP1A2 pool.

```{r table1-ddi}
ddi_safe <- function(caf_mg_day, cip_mg_day, days = 10) {
  ev_for <- function(hep) {
    ev <- rxode2::et(amt = 150, cmt = "stomach_thp", ii = 360,
                     addl = days * 4 - 1)
    if (caf_mg_day > 0) {
      ev <- ev |> rxode2::et(amt = caf_mg_day / 3, cmt = "stomach_caf",
                             ii = 480, addl = days * 3 - 1)
    }
    if (cip_mg_day > 0) {
      ev <- ev |> rxode2::et(amt = cip_mg_day / 2, cmt = "stomach_cip",
                             ii = 720, addl = days * 2 - 1)
    }
    s <- as.data.frame(rxode2::rxSolve(mod,
      ev |> rxode2::et(seq(0, days * 1440, by = 10)), params = covar(0, hep, 1),
      atol = 1e-10, rtol = 1e-8, maxsteps = 1e7))
    max(s$Cthp[s$time >= (days - 1) * 1440])
  }
  f <- function(h) ev_for(h) - 20
  if (f(0.05) < 0) return(NA_real_)
  uniroot(f, c(0.05, 6), tol = 1e-3)$root
}
t1ddi <- data.frame(
  scenario = c("THP alone", "THP + 266 mg CAF", "THP + 960 mg CAF",
               "THP + 500 mg CIP", "THP + 1000 mg CIP", "THP + 1500 mg CIP"),
  caf = c(0, 266, 960, 0, 0, 0), cip = c(0, 0, 0, 500, 1000, 1500),
  published = c(0.62, 0.67, 0.79, 0.70, 0.75, 0.82)
)
t1ddi$reproduced <- round(mapply(ddi_safe, t1ddi$caf, t1ddi$cip), 3)
t1ddi$pct_diff <- round(100 * (t1ddi$reproduced - t1ddi$published) /
                          t1ddi$published, 1)
knitr::kable(
  t1ddi |> select(scenario, published, reproduced, pct_diff) |>
    rename("Scenario (young man, 600 mg/day)" = scenario,
           "Published safe Dmet" = published, "Reproduced" = reproduced,
           "% difference" = pct_diff)
)

stopifnot(
  # All six rows, spanning a 32% range of safe Dmet, reproduce within 5%.
  max(abs(t1ddi$pct_diff)) < 5,
  # Both interactions are monotone in inhibitor dose, as published.
  all(diff(t1ddi$reproduced[1:3]) > 0),
  all(diff(t1ddi$reproduced[c(1, 4, 5, 6)]) > 0)
)
```

Caffeine is not merely a competitor here. Its CYP1A2 demethylation produces
theophylline directly, so a caffeine arm adds theophylline as well as
inhibiting its clearance. The model carries that route explicitly.

```{r caf-to-thp}
caf_only <- single_dose("caf", 960, hours = 48, by_min = 10)
data.frame(
  quantity = c("peak caffeine (mg/L)", "peak paraxanthine (mg/L)",
               "peak theophylline generated from caffeine (mg/L)"),
  value = round(c(max(caf_only$Ccaf), max(caf_only$Cpx),
                  max(caf_only$Cthp)), 3)
)
stopifnot(
  # Caffeine alone must generate measurable theophylline and paraxanthine.
  max(caf_only$Cthp) > 0.05, max(caf_only$Cpx) > 0.5,
  # and stay far below the 40 mg/L caffeine adverse-reaction threshold the
  # paper quotes, even at 960 mg (about twelve espresso shots).
  max(caf_only$Ccaf) < 40
)
```

# Assumptions and deviations

**Blood-to-plasma ratio.** Equations 1 and 4 use the partition coefficient only
through the composite `P(alpha:plasma)/BP`, and `BP` is not reported anywhere in
the paper or the supplement. It is fixed at 1, so the composite equals the
tabulated coefficient. This is not an arbitrary choice: Table 2's rest-of-body
entries are exactly 1.00 for ciprofloxacin and paraxanthine, and exactly the
note-4 adjustment factors (0.8, 0.6) for theophylline and caffeine, which is
what an unadjusted rest-of-body partition of 1 gives when `BP` is 1.

**The `F` = 1 singularity.** Table 3 gives a fraction absorbed of exactly 1 for
theophylline and caffeine, at which the supplement's own `Ka` formula diverges.
The F -> 1 limit is well defined (absorption becomes gastric-emptying-limited,
which is also why the paper says the theophylline peak occurs about an hour
after an oral dose), and `ka_limit_mult` represents it numerically. The
sensitivity chunk above shows the prediction is converged to better than one
part in ten thousand over a tenfold range, so no fitted number is being hidden
here.

**Which reactions inhibit which.** Equation 4's inhibitor index `l` runs over
"the number of inhibitors that interact with `j`". Because Table 3 lists
several reactions per enzyme for the same compound (theophylline alone has
three on CYP1A2), that phrase admits three readings, which differ by a large
factor. All three were implemented and tested against the paper's own
published numbers; two are falsified.

1. *Every other reaction on the enzyme, including the substrate's own parallel
   pathways.* Falsified decisively. It puts a young man on the maximum protocol
   dose at 28.6 mg/L -- above the adverse-reaction threshold -- and gives a safe
   `Dmet` of 1.48 against the published 0.62, contradicting the paper's central
   conclusion that the protocol is safe for young men. It also gives a 200 mg
   half-life of 14.1 h against a measured 8.7 h and a urinary fraction of 14.6%
   against a published 9-12%.
2. *Every other reaction, excluding the substrate's own.* Reproduces all the
   single-compound quantities and the ciprofloxacin interaction, but
   over-counts caffeine, which contributes four CYP1A2 terms instead of one:
   the safe `Dmet` alongside 266 and 960 mg/day of caffeine comes out at 0.94
   and 1.42 against a published 0.67 and 0.79.
3. *One term per competing compound*, using the Km of that compound's dominant
   (highest `Vmax`) pathway on the enzyme as its Ki. This is what is
   implemented. It is the literal reading of Equation 4 -- `l` indexes
   inhibitors, and `U_l` is "the unbound concentration of inhibitor `l`", one
   per inhibitor -- and it reproduces all six interaction rows of Table 1 to
   within 5%.

Ciprofloxacin cannot distinguish readings 2 and 3, because it has exactly one
CYP1A2 reaction and so contributes one term either way; that is why its rows
are reproduced under both. Caffeine, with four, is what separates them.

**The `Vmax` adjustments of supplement notes 6 and 7 are already applied in
Table 3.** Those notes state that the `Vmax` of all theophylline-metabolising
enzymes was uniformly multiplied by 0.9 and of all caffeine-metabolising
enzymes by 9. Applying them again on top of the tabulated values degrades every
comparison in this vignette -- the safe `Dmet` for young men moves from 0.65 to
0.72 against a published 0.62, and the three ciprofloxacin rows of Table 1 move
from 0.72 / 0.77 / 0.81 to 0.82 / 0.88 / 0.93 against a published
0.70 / 0.75 / 0.82. The tabulated values are therefore post-adjustment, exactly
as Table 2's partition coefficients are post-adjustment for note 4. The
tabulated values are used as given.

**Blood-flow topology.** Equation 1 is written generically, with arterial blood
as the inflow to every compartment. A whole-body model cannot be solved that
way: the lung receives venous blood and the liver receives portal blood from the
gut and spleen. The topology used here -- lung fed by venous blood, liver fed by
the hepatic artery plus gut and spleen outflow, everything else fed arterially
-- is the only one consistent with Supplementary Table 1, whose flows balance
exactly under it and under no other assignment. The hepatic artery (487.5 mL/min
in men, 413 in women) is not tabulated and is derived from that balance.

**Absorbed drug enters the gut wall.** The supplement does not say which
compartment receives absorbed drug. It is routed into the perfused gut
compartment, whose venous outflow drains to the liver, so oral doses undergo
first-pass metabolism. The supplement states drug is not metabolised in the gut
itself, which this respects.

**Luminal states carry amounts, not concentrations.** The Yu and Amidon
equations are written in concentration but no luminal volumes are reported.
Every term in that chain is first order, so the amount form is identical up to a
constant and is what conserves mass; the absorption check above confirms the
tabulated fraction absorbed is delivered exactly.

**Compartment count.** The main text describes "20 well-stirred homogenous
compartments". Supplementary Table 1 resolves twelve organs plus arterial and
venous blood, which is fourteen, and the absorption chain adds nine luminal
states per oral drug. The stated count of 20 cannot be reconstructed from the
tabulated physiology; the tables are used as given.

**Table 3's units column.** The `.docx` conversion rotates that column by seven
rows -- its four blocks have sizes 14 / 15 / 6 / 7 where the parameter types
require 7 / 14 / 15 / 6, because the leading dimensionless (blank) cells are
dropped. Units are therefore assigned from the parameter type, not read off that
column. Its reference column is displaced the same way and per-row citations are
consequently not reproduced; the model file cites Table 3 as a whole.

**The elderly rows of Table 1.** The two young rows reproduce to within 6%, but
elderly men come out at 0.49 against a published 0.35 and elderly women at 0.65
against 0.51. The elderly and young rows differ only in dose (400 vs 600 mg/day)
and in the two activity multipliers, so the discrepancy is in how steeply the
safe threshold scales with dose, i.e. in the degree of saturation of the hepatic
term at lower exposure. Defining the threshold on the mean rather than the peak
concentration improves the young rows slightly (0.61 and 0.82) but does not
close the elderly gap. The sex effect within each age stratum survives, and
young women remain the most exposed group and elderly men the least, so the
clinical conclusions the paper draws from the table are unchanged. The full
rank order is not preserved: because the elderly rows are displaced upward,
reproduced elderly women (0.65) tie reproduced young men (0.65) where the paper
separates them (0.51 against 0.62). The same upward displacement is why the
Figure 4 panel above shows the strongest interaction arms crossing 20 mg/L,
where the paper concludes that an average elderly man on 400 mg/day stays below
it. The interaction terms themselves are not the cause -- the six young-men
interaction rows of Table 1 reproduce within 5% -- so the displacement is in how
the saturable hepatic term scales between the 600 and 400 mg/day dose levels.

**Theophylline Cmax after 5 mg/kg.** Reproduced at 9.50 mg/L against a Figure 3
caption value of 10.3, and the caption's three-point spread across an eightfold
change in `Dmet` (11.3 / 10.3 / 10.0) comes out as 10.29 / 9.50 / 8.70 -- the
same weak dependence, displaced about 8% low.

**Race differences are parameterised, not tabulated.** Supplement section 5
scales the adipose volume by a factor Phi taken from Carpenter 2013, which is
not on disk and could not be retrieved. Rather than invent per-race body-fat
fractions, the model exposes `BODYFAT_PCT` directly, so a user supplying a
measured or published body-fat percentage recovers the paper's Figure 5
mechanism. At the reference value for the subject's sex the adipose volume is
exactly the tabulated one.

**No variability.** The source reports neither between-subject variability nor
a residual-error model, so neither is encoded and none was invented. All
variability in this model is deterministic and covariate-driven.
