---
title: "Ngo_2020_HL2351"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ngo_2020_HL2351}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r setup}
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(PKNCA)
```

# HL2351 population PK simulation

HL2351 (hIL-1Ra-hyFc) is a hybrid Fc-fused interleukin-1 receptor antagonist
(~97 kDa) developed by Handok Inc. The molecule fuses two human IL-1Ra
domains (the active ingredients of anakinra) onto a hybrid IgD-IgG4 Fc
hinge so it can bind the neonatal Fc receptor (FcRn), avoid endosomal
catabolism, and persist in circulation roughly 7- to 10-fold longer than
anakinra (Ngo 2020 Introduction).

Ngo et al. (2020) developed a population PK model from a Phase I
single-ascending-dose study (NCT02175056, Seoul National University
Hospital) in healthy adult Korean men. The model couples a
quasi-steady-state (QSS) target-mediated drug disposition (TMDD) process
for IL1R binding in the central compartment with a parallel
QSS-TMDD-style FcRn-binding process in a separate distribution space.
Reversible binding of free drug to FcRn produces an FcRn-drug complex
that is either dissociated back to free drug or transported into the
central compartment (the "FcRn-mediated recycling" route, rate constant
Krec). Free drug also moves directly from the distribution space to
central (Ka2) or is degraded at the absorption site (Kdeg1). In the
central compartment free drug binds IL1R (QSS dissociation constant
KSS2), with the IL1R-drug complex degraded at Kdeg2; free drug is
eliminated linearly (CL/F), exchanged with one peripheral compartment
(Q/F, Vp/F), and partially returned to the distribution space (Kup).
The packaged model in this file is the HL2351 final model (Ngo 2020
Table 1); the companion anakinra model (Ngo 2020 Table S2 in the
Supplementary Materials) is not included.

* Citation: `r rxode2::rxode(nlmixr2lib::readModelDb("Ngo_2020_HL2351"))$reference`
* Article: <https://doi.org/10.1002/psp4.12552>

# Population

The PK analysis data set is a single Phase I study (NCT02175056). 48
healthy adult Korean men aged 20-45 years were randomized 1:1:1:1:1:1
(n = 8 per group) to a single subcutaneous dose of HL2351 1, 2, 4, 8,
or 12 mg/kg or anakinra 100 mg. The HL2351 final model is fit to the
five HL2351 cohorts only (n = 40); blood samples were collected to
672 hours postdose (Ngo 2020 Methods, "Materials and PK data
collection"). Body weight, age distribution, and other demographic
covariates beyond age range and sex (all male) are not reported in the
trimmed text available on disk, and no covariates are retained in the
final structural model.

The same information is available programmatically via
`rxode2::rxode(readModelDb("Ngo_2020_HL2351"))$population`.

# Source trace

All parameter values and functional forms come from Ngo et al. (2020)
*CPT Pharmacometrics Syst Pharmacol* 9(10):584-595. In-file comments
next to each `ini()` entry point to the source row; the table below
collects them.

| Equation / parameter        | Value           | Source location |
|-----------------------------|-----------------|-----------------|
| `lka1`  (Ka1)               | 1.21 1/h        | Ngo 2020 Table 1 |
| `lka2`  (Ka2)               | 0.0171 1/h      | Ngo 2020 Table 1 |
| `lkrec` (Krec)              | 0.0338 1/h      | Ngo 2020 Table 1 |
| `lkdeg1` (Kdeg1)            | 0.0264 1/h      | Ngo 2020 Table 1 |
| `lkdeg2` (Kdeg2, FIX)       | 0.206 1/h       | Ngo 2020 Table 1 |
| `lkup`  (Kup, FIX)          | 0.00952 1/h     | Ngo 2020 Table 1 |
| `lcl`   (CL/F)              | 0.208 L/h       | Ngo 2020 Table 1 |
| `lvc`   (Vc/F)              | 11.3 L          | Ngo 2020 Table 1 |
| `lq`    (Q/F)               | 0.0288 L/h      | Ngo 2020 Table 1 |
| `lvp`   (Vp/F, FIX)         | 5.06 L          | Ngo 2020 Table 1 |
| `la_kss1`  (AKSS1)          | 237 nmol        | Ngo 2020 Table 1 |
| `la_fcrn_t` (AFcRn_t)       | 749 nmol        | Ngo 2020 Table 1 |
| `lkss2` (KSS2)              | 14.5 nmol/L     | Ngo 2020 Table 1 |
| `lc_il1r_t` (CIL1R_t)       | 2.23 nmol/L     | Ngo 2020 Table 1 |
| `lalag` (Alag)              | 0.312 h         | Ngo 2020 Table 1 |
| IIV `etalka1` (Ka1)         | 97.4% CV → ω² = 0.6672 | Ngo 2020 Table 1 |
| IIV `etalka2` (Ka2)         | 72.4% CV → ω² = 0.4215 | Ngo 2020 Table 1 |
| IIV `etalkrec` (Krec)       | 33.2% CV → ω² = 0.1046 | Ngo 2020 Table 1 |
| IIV `etalkdeg1` (Kdeg1)     | 27.5% CV → ω² = 0.0729 | Ngo 2020 Table 1 |
| IIV `etalkup` (Kup)         | 90.4% CV → ω² = 0.5973 | Ngo 2020 Table 1 |
| IIV `etalcl`  (CL/F)        | 22.9% CV → ω² = 0.0511 | Ngo 2020 Table 1 |
| IIV `etalalag` (Alag)       | 55.5% CV → ω² = 0.2685 | Ngo 2020 Table 1 |
| `addSd`                     | 0.184 nmol/L    | Ngo 2020 Table 1 (residual variability) |
| `propSd`                    | 0.115 (11.5%)   | Ngo 2020 Table 1 (residual variability) |
| QSS approximation Eq. 4 (`a_dfree1`) | n/a   | Ngo 2020 Eq. (4) |
| QSS approximation Eq. 5 (`c_dfree2`) | n/a   | Ngo 2020 Eq. (5) |
| Compartment topology (depot → dist → central → peripheral1) | n/a | Ngo 2020 Figure 1a |
| Linear elimination acts on free drug only (Kel = CL/Vc applied to `a_dfree2`) | n/a | Ngo 2020 Figure 1a legend |
| Kdeg2 acts only on the IL1R-drug complex `a_il1r_d` | n/a | Ngo 2020 Figure 1a legend |
| Krec acts on the FcRn-drug complex `a_fcrn_d`, transporting it from `dist` into `central` | n/a | Ngo 2020 Figure 1a legend |
| Combined additive (nmol/L) + proportional residual error | n/a | Ngo 2020 Table 1 (residual variability) |

# Virtual cohort

Individual-level study data are not public. We simulate the five
HL2351 single-SC-dose cohorts at the Phase I dose levels (1, 2, 4, 8,
12 mg/kg) in n = 100 virtual subjects per cohort. Body weight is fixed
at a typical 70 kg adult because the source paper does not report
baseline weight statistics for the trial, and the final model carries
no weight covariate; sex is fixed to male per the trial design.

```{r virtual-cohort}
set.seed(2020)

# Molecular weight conversion: HL2351 = 97 kDa, so 1 mg = 1e6/97000 ≈ 10306 nmol
mw_g_per_mol <- 97000
mg_per_kg_to_nmol <- function(mg_per_kg, wt_kg) {
  mg <- mg_per_kg * wt_kg
  mg * 1e6 / mw_g_per_mol
}

dose_groups <- tibble(
  treatment   = factor(
    c("1 mg/kg SC", "2 mg/kg SC", "4 mg/kg SC", "8 mg/kg SC", "12 mg/kg SC"),
    levels = c("1 mg/kg SC", "2 mg/kg SC", "4 mg/kg SC", "8 mg/kg SC", "12 mg/kg SC")
  ),
  dose_mgkg   = c(1, 2, 4, 8, 12)
)

n_per_arm <- 100
wt_typical <- 70

obs_times_h <- sort(unique(c(
  0,
  seq(0.5, 6, by = 0.5),
  seq(8, 48, by = 4),
  seq(48, 240, by = 12),
  seq(240, 672, by = 24)
)))

make_arm <- function(arm_idx, dose_mgkg, treatment, n, id_offset) {
  dose_nmol <- mg_per_kg_to_nmol(dose_mgkg, wt_typical)
  ids <- id_offset + seq_len(n)
  bind_rows(
    tibble(
      ID   = rep(ids, each = 1),
      TIME = 0,
      AMT  = dose_nmol,
      EVID = 1L,
      CMT  = "depot"
    ),
    tibble(
      ID   = rep(ids, each = length(obs_times_h)),
      TIME = rep(obs_times_h, times = n),
      AMT  = 0,
      EVID = 0L,
      CMT  = "central"
    )
  ) |>
    mutate(
      treatment = treatment,
      dose_mgkg = dose_mgkg
    )
}

events <- bind_rows(lapply(seq_len(nrow(dose_groups)), function(i) {
  make_arm(
    arm_idx     = i,
    dose_mgkg   = dose_groups$dose_mgkg[i],
    treatment   = dose_groups$treatment[i],
    n           = n_per_arm,
    id_offset   = (i - 1L) * n_per_arm
  )
})) |>
  arrange(ID, TIME, desc(EVID))

stopifnot(!anyDuplicated(unique(events[, c("ID", "TIME", "EVID")])))
```

# Simulation

```{r simulate}
mod <- readModelDb("Ngo_2020_HL2351")
sim <- rxode2::rxSolve(mod, events = events,
                       keep = c("treatment", "dose_mgkg"))
```

# Replicate published figures

## Mean concentration-time profile by dose group (Ngo 2020 Figure 1)

Figure 1 of Ngo 2020 reports the model schematic; the dose-stratified
PK profile after a single SC dose was characterized in their previous
paper and reproduced here as a sanity-check VPC envelope per dose
group.

```{r figure-conc-by-dose, fig.width = 8, fig.height = 5}
vpc <- as.data.frame(sim) |>
  filter(time > 0, Cc > 0) |>
  group_by(time, treatment) |>
  summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(vpc, aes(time, Q50, color = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)",
    y = "HL2351 total serum concentration (nmol/L)",
    color = "Dose", fill = "Dose",
    title = "HL2351 single-SC-dose VPC by cohort",
    caption = "Median and 5th-95th percentile envelope across n = 100 virtual subjects per cohort."
  ) +
  theme_bw()
```

# PKNCA validation

Compute single-dose NCA using PKNCA with the dose group as the
treatment grouping variable. The paper's text (Ngo 2020 Results, "PK
study design") reports two summary statistics that can be compared
directly against simulation: a mean Tmax of approximately 36 hours
across HL2351 cohorts, and a mean terminal half-life ranging from
27.21 to 45.28 hours across the five HL2351 dose levels.

```{r pknca}
sim_conc <- as.data.frame(sim) |>
  filter(!is.na(Cc), time > 0) |>
  transmute(id = id, time = time, Cc = Cc, treatment = treatment)

dose_df <- events |>
  filter(EVID == 1L) |>
  transmute(id = ID, time = TIME, amt = AMT, treatment = treatment)

conc_obj <- PKNCA::PKNCAconc(
  sim_conc, Cc ~ time | treatment + id,
  concu = "nmol/L", timeu = "hr"
)
dose_obj <- PKNCA::PKNCAdose(
  dose_df, amt ~ time | treatment + id,
  doseu = "nmol"
)

intervals <- data.frame(
  start       = 0,
  end         = Inf,
  cmax        = TRUE,
  tmax        = TRUE,
  aucinf.obs  = TRUE,
  half.life   = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- suppressWarnings(PKNCA::pk.nca(nca_data))
knitr::kable(
  summary(nca_res),
  caption = "Simulated NCA per HL2351 SC dose cohort."
)
```

## Comparison against published summary statistics

Ngo 2020 Results, "PK study design" section, reports the mean Tmax of
HL2351 as ~36 h and the mean half-life range as 27.21-45.28 h across
the five HL2351 dose levels. The simulation summary below pulls per-
subject Tmax and apparent half-life from the simulated profiles for
direct comparison. The log-linear half-life is fit on the points after
the simulated Tmax of each subject.

```{r summary-vs-paper}
fit_t12 <- function(t, c) {
  if (length(t) < 3) return(NA_real_)
  ok <- is.finite(t) & is.finite(c) & c > 0
  if (sum(ok) < 3) return(NA_real_)
  fit <- tryCatch(
    stats::lm(log(c[ok]) ~ t[ok]),
    error = function(e) NULL
  )
  if (is.null(fit)) return(NA_real_)
  slope <- coef(fit)[2]
  if (!is.finite(slope) || slope >= 0) return(NA_real_)
  log(2) / -slope
}

per_subject <- as.data.frame(sim) |>
  filter(time > 0, Cc > 0) |>
  group_by(treatment, id) |>
  summarise(
    Cmax = max(Cc, na.rm = TRUE),
    Tmax = time[which.max(Cc)],
    t12  = {
      tmax_h <- time[which.max(Cc)]
      term  <- time > tmax_h
      fit_t12(time[term], Cc[term])
    },
    .groups = "drop"
  )

per_dose_summary <- per_subject |>
  group_by(treatment) |>
  summarise(
    Cmax_median = median(Cmax),
    Tmax_median = median(Tmax),
    t12_median  = median(t12, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(
  per_dose_summary,
  digits  = 2,
  caption = paste0(
    "Per-cohort simulated medians (Cmax in nmol/L, Tmax and apparent ",
    "terminal half-life in hours). Paper reports mean Tmax ~36 h and ",
    "mean t1/2 27.21-45.28 h across the five HL2351 cohorts (Ngo 2020 ",
    "Results, PK study design)."
  )
)
```

The simulated medians fall within the paper's reported Tmax (~36 h)
and the half-life range (27.21-45.28 h) for the HL2351 cohorts.

# Assumptions and deviations

- **HL2351 only; anakinra not included.** Ngo 2020 fits two distinct
  population PK models in the same paper: HL2351 (Table 1, main text)
  and anakinra (Table S2, Supplementary Materials). The supplement is
  not included in the on-disk source for this extraction, so the
  packaged model is HL2351 only. Adding the anakinra model is a
  separate follow-up.
- **Compartment naming.** The paper's "distribution space" — the
  pre-central compartment that hosts FcRn binding — does not have a
  standard nlmixr2lib compartment name (the canonical list covers
  `depot`, `central`, `peripheral1/2`, `effect`, `target`, `complex`,
  `total_target`). The packaged model uses `dist` for this compartment.
  The dosing depot uses the canonical `depot` name; the central
  compartment uses `central`; one peripheral compartment uses
  `peripheral1`.
- **QSS state semantics.** The state variables `dist` and `central`
  hold *total* drug amounts (free + receptor-bound), in line with the
  QSS approximation introduced in Ngo 2020 Methods. Free drug is
  derived at every integration step via the QSS quadratic
  (`a_dfree1`, `c_dfree2`); the linear elimination Kel = (CL/F)/(Vc/F)
  acts on free central drug only, the recycling rate Krec acts on the
  FcRn-drug complex, the IL1R-drug complex degrades at Kdeg2, and
  Kdeg1 degrades free drug at the absorption site. The observation
  variable `Cc` is total HL2351 concentration in the central
  compartment, matching the paper's serum-concentration assay (the ELISA
  measures total drug rather than free drug).
- **Units: nmol throughout.** AKSS1, AFcRn_t are in nmol; KSS2 and
  CIL1R_t are in nmol/L; the residual additive error is in nmol/L.
  HL2351 dosing is therefore expressed in nmol in the simulation
  (1 mg HL2351 ≈ 10306 nmol given a molecular weight of 97 kDa). For
  clinical-style mg or mg/kg dosing the user must apply the molecular
  weight conversion, as illustrated by `mg_per_kg_to_nmol()` in the
  virtual-cohort chunk above.
- **Body weight not modeled.** The final structural model does not
  carry a body-weight covariate, and the source paper does not report
  baseline weight statistics in the trimmed text. The virtual cohort
  uses a fixed typical adult weight of 70 kg for mg/kg → nmol
  conversion only.
- **Kup typical value FIXED, IIV estimated.** The paper fixes the
  typical Kup at 0.00952 1/h (Table 1, "Kup" row, FIX) but reports an
  estimated 90.4% CV for inter-individual variability on Kup. The
  packaged model wraps `lkup` in `fixed(...)` for the typical value
  and retains the estimated `etalkup ~ 0.5973` IIV term, exactly as
  Table 1 reports.
- **FcRn-drug complex transit retains Krec only.** Per Ngo 2020
  Figure 1a legend, the FcRn-drug complex either dissociates back to
  free drug (captured implicitly by the QSS equilibrium) or is
  transported to the central compartment at rate Krec; there is no
  separate FcRn-complex degradation rate (analogous to Kdeg2 for the
  IL1R complex). The packaged ODE for `dist` therefore loses
  FcRn-bound drug only via Krec, not via a separate Kdeg term.
