---
title: "Levofloxacin (Nguyen 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Levofloxacin (Nguyen 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Nguyen_2026_levofloxacin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Nguyen_2026_levofloxacin"))$description`
* Article: <https://doi.org/10.1007/s40262-026-01619-3>
* Supplement (Supplementary Information S1-S5, Tables S1-S4, Figures S1-S8):
  <https://doi.org/10.1007/s40262-026-01619-3> (open access; also retrievable
  from EuropePMC as `PMC13038660`)

Levofloxacin is a core drug in multidrug-resistant tuberculosis (MDR-TB)
regimens, and its 24-hour exposure `AUC(0-24)` is the metric that drives both
bactericidal activity and the suppression of acquired resistance. Therapeutic
drug monitoring is therefore recommended for high-risk patients, but the
conventional route to `AUC(0-24)` is intensive blood sampling, which is
impractical in the community and high-burden settings where most MDR-TB care
happens. Nguyen 2026 fits a joint plasma + saliva population PK model so that a
plasma `AUC(0-24)` can be estimated from saliva alone, then shows that a
*single* saliva sample at 2 h post-dose predicts it with a mean difference of
under 1%.

This is the levofloxacin member of a pair. The same group published a
structurally similar linezolid model in the same journal and year, packaged here
as `modellib("Nguyen_2026_linezolid")`; the two differ in an important
structural detail noted under *Model structure* below.

# Population

The model was fitted to 342 paired plasma-saliva samples from 57 adults with
MDR-TB (Nguyen 2026 Table 1), a pharmacokinetic sub-study of the V-SMART trial
(ACTRN12620000681954) run across four provinces in Vietnam. Sixty patients met
the inclusion criteria; three who had completed their levofloxacin course had
below-limit-of-quantification samples and were excluded, leaving 57 with
evaluable data.

Median (IQR) age was 44 (33-51) years, weight 50 (45-56) kg, height 162
(160-167) cm, fat-free mass 41.4 (36.7-45.6) kg and BMI 19.1 (17.5-20.8)
kg/m^2 -- a lean cohort by international standards. Forty patients (70.2%) were
male. All had renal and hepatic function markers within normal limits
(creatinine clearance 71 (65.1-84.7) mL/min, serum creatinine 77 (71-85) umol/L,
ALT 20 (15-36) U/L, AST 24 (19-32) U/L), so the model carries no information
about organ impairment -- which matters, because levofloxacin is predominantly
renally cleared.

Levofloxacin was given orally at the WHO-recommended 750-1000 mg once daily,
equivalent to 15-20 mg/kg/day (median 17.9 mg/kg/day). Sampling was pre-dose,
2 h and 5 h post-dose after at least 7 days of treatment, so every observation
is a steady-state observation. The assay lower limit of quantification was
0.5 mg/L in both matrices; one plasma value (0.49 mg/L) and its paired saliva
value (0.37 mg/L) fell below it and were retained "as measured" rather than
censored.

The same information is available programmatically from the model's `population`
metadata:

```{r population}
pop <- rxode2::rxode(readModelDb("Nguyen_2026_levofloxacin"))$population
str(pop, max.level = 1)
```

# Model structure

The plasma model is one-compartment with first-order absorption, an absorption
lag time and linear elimination. A two-compartment structure and a transit
absorption chain were both investigated and rejected; a three-compartment model
was not attempted given the sparse design (Methods 2.3, Table S1). Saliva is
added as a **kinetically distinct compartment**, which fitted better than
rescaling the plasma concentration by a saliva:plasma scale factor (objective
function value 695.794 versus 700.743, Table 2 versus Table S2).

That is the same choice the group made for linezolid, and the opposite of the
choice made for busulfan in `modellib("Xu_2023_busulfan")`, where a scale factor
(`lfsaliva = log(0.88)`) beat a distinct saliva compartment by `dOFV = -82.52`.
The two structures are genuine competitors and the answer is drug-specific, so
nlmixr2lib carries both encodings.

```{r structure}
cat(rxode2::rxode(readModelDb("Nguyen_2026_levofloxacin"))$modelDesc)
```

Two structural points distinguish this model from its linezolid sibling and are
worth stating explicitly, because both become exact validation gates further
down.

**One saliva leg, not two.** The linezolid model carries three saliva rate
constants -- secretion into saliva, reabsorption back towards plasma, and
irreversible salivary loss. This levofloxacin model carries only two. Figure 1
draws a single dotted arrow from the central compartment into the saliva
compartment (`Kabs`) and a single dotted arrow out of it (`Kel`), with no return
path, and the Table S4 `$PK` block defines only `K23` and `K30` -- there is no
`K32`.

**The saliva state is driven, not mass-balance-coupled.** Drug appears in saliva
in proportion to the central amount, but the central compartment is not depleted
by it. Methods 2.3 says so in as many words: the plasma model was "extended to
include a saliva bio-compartment (i.e., a hypothetical effect compartment, which
does not account for mass balance)". See *Assumptions and deviations* below for
the arithmetic that confirms it, because the supplementary control stream
(Table S4) is abridged and does not print the `$DES` block.

Together these give two identities that hold subject by subject:

1. Total elimination is the plasma clearance alone, so at steady state
   `AUC(0-24) = Dose / (CL/F)` exactly.
2. Integrating the saliva equation over a steady-state dosing interval gives
   `kin_saliva * AUC_central = kel_saliva * AUC_saliva`, so the saliva:plasma
   exposure ratio is the **parameter-free constant**
   `kin_saliva / kel_saliva`, independent of dose, clearance and volume.

# Source trace

Per-parameter origins are recorded as in-file comments beside each `ini()` entry
in `inst/modeldb/specificDrugs/Nguyen_2026_levofloxacin.R`. They are collected
here for review. All values are the **final model** column of Nguyen 2026
Table 2 (objective function value = 695.794). Table S1 (base plasma model) and
Table S2 (competing scale-factor saliva model) are rejected alternatives and are
not used for any value.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL/F) | 10.311 L/h | Table 2 final model `theta_CL` (RSE 0.01%; SIR median 10.310, 95% CI 10.308-10.312); Table S4 `$THETA (8, 10.3, 12)` |
| `lvc` (V/F) | 278.88 L | Table 2 final model `theta_V` (RSE 0.008%; SIR median 278.88, 95% CI 278.82-278.92); Table S4 `$THETA (250, 278, 300)` |
| `lka` (Ka) | 4.18 1/h, **fixed** | Table 2 final model "4.18 (Fixed)"; Table S4 `$THETA (4.18) FIX`. Results 3.2: fixed "based on published estimates to stabilise the model estimates in terms of convergence and gradients", because few concentrations were observed in the absorption phase; the no-lag base model estimated 2.77 1/h with %RSE 105% (Table S1) |
| `ltlag` (Tlag) | 0.95 h, **fixed** | Table 2 final model "0.95 (Fixed)"; Table S4 `$THETA (0.95) FIX`. Estimated at 0.947 h with RSE 3% in the base plasma model (Table S1), but %RSE rose to 141% in the combined plasma-plus-saliva fit, so it was held at the rounded value (Results 3.2) |
| `lkin_saliva` (Kabs) | 4.929 1/h | Table 2 final model `K_abs` (RSE 0.01%; SIR median 4.929, 95% CI 4.927-4.930); Table S4 `K23`, `$THETA (1, 4.94, 6)`; Figure 1 "first-order saliva absorption rate" |
| `lkel_saliva` (Kel) | 5.084 1/h | Table 2 final model `K_el` (RSE 0.02%; SIR median 5.084, 95% CI 5.082-5.086); Table S4 `K30`, `$THETA (1, 5.08, 6)`; Figure 1 "elimination rate from saliva compartment" |
| Saliva reabsorption leg | **absent** | Figure 1 draws no return arrow; Table S4 `$PK` defines `K23` and `K30` only, with no `K32`. Contrast `modellib("Nguyen_2026_linezolid")`, which has all three |
| Covariate effects | **none** | Results 3.2. Allometric body weight: `dOFV = -1.93` (fixed exponents), `-0.64` (estimated). Allometric fat-free mass: `dOFV = +5.42` (fixed), `+4.9` (estimated). Age, sex, renal and hepatic markers: "none of the other covariates significantly correlated with clearance or volume (Fig. S8)". Forward inclusion required `dOFV <= -3.84` |
| `etalcl` | variance 0.169744 | Table 2 final model `omega_CL` = 41.2 CV% (RSE 62.9%, shrinkage 2%). Variance convention pinned below |
| `etalvc` | variance 0.485809 | Table 2 final model `omega_V` = 69.7 CV% (RSE 11.1%, shrinkage 12%). Diagonal `$OMEGA`, no `BLOCK`, so CL and V are uncorrelated |
| IIV on Ka, Tlag, Kabs, Kel | none | Table S4 `$OMEGA` "0 FIX" on all four; Results 3.2: IIV on `Kabs` and `Kel` "could not be estimated because including them rendered the model unstable" |
| `propSd` (plasma) | 0.224 | Table 2 final model `sigma_Plasma_Prop` = 22.4 CV% (RSE 0.5%, shrinkage 19%) |
| `addSd` (plasma) | 0.1394 mg/L | Table 2 final model `sigma_Plasma_Add` (RSE 0.02%, shrinkage 19%) |
| `propSd_Csaliva` (saliva) | 0.3453 | Table 2 final model `sigma_Saliva_Prop` = 34.53 CV% (RSE 0.2%, shrinkage 8%) |
| `addSd_Csaliva` (saliva) | 0.06 mg/L | Table 2 final model `sigma_Saliva_Add` (RSE 0.03%, shrinkage 8%) |
| `d/dt(depot)`, `d/dt(central)` | n/a | Methods 2.3; Table S4 `$MODEL COMP = (ABS)`, `COMP = (CENTRAL)`, `$PK K20 = CL / V` |
| `alag(depot)` | n/a | Table S4 `$PK ALAG1 = TVALAG*EXP(ETA(4))`; compartment 1 is `COMP = (ABS)` |
| `d/dt(saliva)` | n/a | Figure 1 conceptual model; Table S4 `COMP = (SALIVA)`, `K23`/`K30`. `$DES` is not printed -- reconstruction documented below |
| `Cc <- central / vc` | n/a | Table S4 `$ERROR CP = A(2)/V`, `$PK S2 = V` |
| `Csaliva <- saliva / vc` | n/a | Reconstructed; Table S4 prints `CS = A(3)` -- see below |
| Exponential IIV form | n/a | Supplementary Information S1.1 Eq. 1, `Pi = PTV * exp(eta_i)` |
| Combined residual form | n/a | Supplementary Information S1.2 Eq. 4, `Y = IPRED * (1 + EPS(1)) + EPS(2)` |

## The variance convention, pinned exactly

Table 2 labels its variability rows "CV%", which is ambiguous: for a
log-normally distributed parameter `CV%` may mean `sqrt(variance) * 100` or
`sqrt(exp(variance) - 1) * 100`, and at this magnitude the two differ by nearly
nine percentage points on the volume row. Table S4 settles it.

The control stream's `$THETA` initials are the final estimates rounded -- 10.3,
278, 4.94 and 5.08 against final estimates of 10.311, 278.88, 4.929 and 5.084 --
so the run was seeded from the fit it reports, and its `$OMEGA` and `$SIGMA`
initials are the final variances to the same rounding.

```{r variance-convention}
tibble::tribble(
  ~term,                 ~S4_initial, ~reported, ~reported_unit,
  "omega_CL",                  0.166,     41.2,  "CV%",
  "omega_V",                   0.48,      69.7,  "CV%",
  "sigma_Plasma_Prop",         0.0501,    22.4,  "CV%",
  "sigma_Plasma_Add",          0.0195,     0.1394, "mg/L",
  "sigma_Saliva_Prop",         0.1240,    34.53, "CV%",
  "sigma_Saliva_Add",          0.00364,    0.06, "mg/L"
) |>
  mutate(
    as_sd    = sqrt(S4_initial) * if_else(reported_unit == "CV%", 100, 1),
    as_lnorm = if_else(
      reported_unit == "CV%",
      sqrt(exp(S4_initial) - 1) * 100,
      NA_real_
    )
  ) |>
  dplyr::rename(
    "Term"                   = term,
    "Table S4 initial"       = S4_initial,
    "Table 2 value"          = reported,
    "Unit"                   = reported_unit,
    "sqrt(var)"              = as_sd,
    "sqrt(exp(var)-1)*100"   = as_lnorm
  ) |>
  knitr::kable(
    digits  = 4,
    caption = paste(
      "Table S4's initials against Table 2's reported values. The reported",
      "numbers match sqrt(variance), not the log-normal back-transform."
    )
  )
```

The two **additive** rows are what make this airtight, because an additive
residual error is a concentration in mg/L and no CV convention applies to it at
all: `sqrt(0.0195) = 0.1396` against a reported `0.1394 mg/L`, and
`sqrt(0.00364) = 0.0603` against a reported `0.06 mg/L`. The control stream
holds variances and the table prints standard deviations. Applying the same
reading to the two IIV rows gives `sqrt(0.166) = 40.74%` against 41.2% and
`sqrt(0.48) = 69.28%` against 69.7%; the log-normal reading would have required
78.49% on the volume row, which is not what Table 2 prints. The model file
therefore encodes `etalcl ~ 0.412^2` and `etalvc ~ 0.697^2`.

# Virtual cohort

Original observed data are not publicly available, and the paper does not
publish per-subject doses -- only the 750-1000 mg once-daily range and the
17.9 mg/kg/day median. The cohort below simulates the two endpoints of that
range as separate arms rather than guessing a mix. No covariate enters the
model, so the weight drawn below is carried for reporting only; it does not
affect any simulated concentration.

```{r cohort}
# `set.seed()` seeds R's RNG, which is what draws the weights below. It does NOT
# seed rxode2's simulation RNG (that is `rxSetSeed()`, called before each solve),
# and rxode2's streams are partitioned PER SOLVER THREAD -- so the etas drawn
# below differ between a 2-core CI runner and a 16-thread workstation and no seed
# can make them agree. Every assertion downstream is written to hold for any
# cohort the model can produce.
set.seed(20260912)

n_per_arm <- 200            # skill cap is 200 per arm
tau       <- 24             # dosing interval (h)

# Doses are Nguyen 2026 Results 3.1: "the World Health Organisation (WHO)
# recommended dose of 750-1000 mg once daily".
doses <- c("750 mg q24h" = 750, "1000 mg q24h" = 1000)

# Number of daily doses needed for steady state. With no IIV on Ka or Tlag, the
# terminal half-life is ln2 * V / CL, and log(V/CL) has SD
# sqrt(0.412^2 + 0.697^2) = 0.81 with a median half-life of ln2*278.88/10.311 =
# 18.7 h. The extreme subject in a 400-draw cohort sits about 3.2 SD out, i.e.
# a half-life near 250 h, so the STOCHASTIC cohort's dose train has to be long
# enough for THAT subject and not just for the typical one. 121 doses is 2880 h,
# about 11 half-lives for the extreme case. Verified below rather than assumed.
#
# The deterministic typical-value block further down uses a much shorter train
# (21 doses = 480 h, over 25 half-lives at the typical value) because it draws
# no etas and so has no slow tail to wait for. That is not an optimisation: a
# 2880 h train combined with the 0.02 h output grid that block needs exceeds
# rxode2's solver step budget and returns "could not solve the system".
n_dose_cohort  <- 121
t_last         <- tau * (n_dose_cohort - 1)
n_dose_typical <- 21
t_last_typical <- tau * (n_dose_typical - 1)

log_sd_thalf <- sqrt(0.412^2 + 0.697^2)
c(median_half_life_h  = log(2) * 278.88 / 10.311,
  log_sd_half_life    = log_sd_thalf,
  extreme_half_life_h = log(2) * 278.88 / 10.311 * exp(3.2 * log_sd_thalf),
  cohort_train_h      = t_last,
  typical_train_h     = t_last_typical)

# Table 1 gives weight as median 50 kg, IQR 45-56 kg. For a log-normal,
# log(56) - log(45) spans 2 * qnorm(0.75) standard deviations.
wt_sd_log <- (log(56) - log(45)) / (2 * stats::qnorm(0.75))

# `wt = NULL` draws weight from the cohort distribution; passing a number pins
# it, which the typical-value block below uses.
make_cohort <- function(n, dose, label, id_offset = 0L, obs_times, wt = NULL,
                        n_dose = n_dose_cohort) {
  last <- tau * (n_dose - 1)
  subj <- tibble::tibble(
    id        = id_offset + seq_len(n),
    WT        = if (is.null(wt)) stats::rlnorm(n, log(50), wt_sd_log) else wt,
    treatment = label
  )
  dosing <- subj |>
    tidyr::crossing(time = seq(0, last, by = tau)) |>
    mutate(amt = dose, evid = 1L, cmt = "depot", dvid = NA_integer_)
  obs <- subj |>
    tidyr::crossing(time = last + obs_times) |>
    # Observation rows sit on the ODE STATE `central`, never on the algebraic
    # observable `Cc`: naming an observable in `cmt` injects a compartment slot
    # for it after the ODE states and renumbers everything. rxode2 returns both
    # `Cc` and `Csaliva` as columns at these rows.
    mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L)
  bind_rows(dosing, obs) |> arrange(id, time, desc(evid))
}

# The paper reconstructed 25-point plasma profiles "representing hourly sampling
# from 0 to 24 hours" (Supplementary Information S4.1), so the NCA cohort
# reproduces that measurement grid rather than improving on it. One extra record
# one interval past the last dose supports the steady-state check below.
grid_hourly <- 0:24

events <- bind_rows(
  make_cohort(n_per_arm, doses[["750 mg q24h"]], "750 mg q24h",
              id_offset = 0L, obs_times = grid_hourly),
  make_cohort(n_per_arm, doses[["1000 mg q24h"]], "1000 mg q24h",
              id_offset = as.integer(n_per_arm), obs_times = grid_hourly)
)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

Steady state is verified rather than assumed. There is no dose at
`t_last + tau`, so the concentration there is the trough that would precede the
next dose; comparing it with the trough at `t_last` measures any residual
accumulation directly.

```{r steady-state-check}
mod <- readModelDb("Nguyen_2026_levofloxacin")

rxode2::rxSetSeed(4321)
ss <- rxode2::rxSolve(
  mod,
  events    = events |> filter(treatment == "750 mg q24h"),
  keep      = c("treatment", "WT"),
  useLinCmt = FALSE      # rxode2's ODE->linCmt auto-conversion breaks the
) |>                     # dvid mapping for multi-output models
  as.data.frame() |>
  filter(time %in% c(t_last, t_last + tau)) |>
  select(id, time, Cc) |>
  tidyr::pivot_wider(names_from = time, values_from = Cc, names_prefix = "t")

ss_rel <- abs(ss[[paste0("t", t_last + tau)]] / ss[[paste0("t", t_last)]] - 1)
summary(ss_rel)

# Accumulation is complete to well under 1% for every subject, including the
# slowest the model can draw. 2% leaves headroom over the solver tolerance while
# still going red if the dose train were cut to a handful of doses.
stopifnot(max(ss_rel) < 0.02)
```

# Simulation

One `rxSolve()` call per arm: solving an `rxUi` is quadratic in the number of
subjects per call, so splitting the arms is materially cheaper than one combined
solve. Each arm is seeded separately so the two arms are independent draws
rather than sharing a stream.

```{r simulate}
solve_arm <- function(arm, seed) {
  rxode2::rxSetSeed(seed)
  rxode2::rxSolve(
    mod,
    events    = events |> filter(treatment == arm),
    keep      = c("treatment", "WT"),
    useLinCmt = FALSE
  ) |>
    as.data.frame()
}

sim <- bind_rows(
  solve_arm("750 mg q24h", 10750),
  solve_arm("1000 mg q24h", 11000)
) |>
  # Re-time so t = 0 is the last dose. PKNCA then sees a clean 0-24 h
  # steady-state interval with a genuine time-zero (trough) record.
  mutate(tad = time - t_last) |>
  filter(tad >= 0, tad <= tau)

stopifnot(!anyNA(sim$Cc), !anyNA(sim$Csaliva), all(sim$Cc >= 0))
# Fail loudly if the grid ever stops producing the time-zero record PKNCA
# anchors AUC(0-tau) on, rather than silently injecting Cc = 0 -- at steady
# state the pre-dose concentration is the trough, NOT zero.
stopifnot(all(table(sim$id[sim$tad == 0]) == 1L))
```

# Exact closed-form gates

Both identities in *Model structure* are properties of the ODE system, not of
the drawn cohort: they hold subject by subject with the subject's own drawn
parameters, so the only discrepancy is trapezoidal-integration error. That makes
a tight bound the correct assertion here (unlike the cohort-derived comparisons
further down, where robust statistics are used instead).

Critically, each gate is stated against a **literal transcribed from Table 2**
-- `10.311 L/h`, `278.88 L`, `4.929 1/h`, `5.084 1/h` -- not against the model's
own parameter variables. A gate built from `first(cl)` would agree with itself
no matter what number the model file carried; these go red on a transcription
error.

```{r closed-form}
# Published Table 2 point estimates, hardcoded. These are the comparator.
pub_cl   <- 10.311   # L/h
pub_vc   <- 278.88   # L
pub_kabs <- 4.929    # 1/h
pub_kel  <- 5.084    # 1/h

# A single typical-value subject per arm on a dense grid, so the trapezoid is
# not the limiting error. zeroRe() removes the IIV, making these rows fully
# deterministic (no covariate enters the model, so nothing else varies), which
# is also why the shorter 21-dose train suffices here.
dense <- seq(0, tau, by = 0.02)

cf_events <- bind_rows(
  make_cohort(1, doses[["750 mg q24h"]], "750 mg q24h", id_offset = 0L,
              obs_times = dense, wt = 50, n_dose = n_dose_typical),
  make_cohort(1, doses[["1000 mg q24h"]], "1000 mg q24h", id_offset = 1L,
              obs_times = dense, wt = 50, n_dose = n_dose_typical)
)

cf <- rxode2::rxSolve(
  rxode2::zeroRe(mod), events = cf_events,
  keep = c("treatment", "WT"), useLinCmt = FALSE
) |>
  as.data.frame() |>
  mutate(tad = time - t_last_typical) |>
  filter(tad >= 0, tad <= tau)

stopifnot(nrow(cf) == 2L * length(dense), !anyNA(cf$Cc))

# The shorter train must still have reached steady state, or every gate below
# is measuring an accumulating profile. At steady state the concentration one
# full interval after the last dose equals the trough that preceded it.
ss_typical <- cf |>
  filter(tad %in% c(0, tau)) |>
  group_by(treatment) |>
  summarise(rel = abs(Cc[tad == tau] / Cc[tad == 0] - 1), .groups = "drop")
ss_typical
stopifnot(max(ss_typical$rel) < 1e-3)

trap <- function(x, y) sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)

cf_res <- cf |>
  group_by(id, treatment) |>
  summarise(
    dose     = doses[[first(treatment)]],
    auc_pl   = trap(tad, Cc),
    auc_sal  = trap(tad, Csaliva),
    .groups  = "drop"
  ) |>
  mutate(
    auc_published     = dose / pub_cl,
    auc_pct_err       = 100 * (auc_pl / auc_published - 1),
    ratio_observed    = auc_sal / auc_pl,
    ratio_published   = pub_kabs / pub_kel,
    ratio_pct_err     = 100 * (ratio_observed / ratio_published - 1)
  )

cf_res |>
  select(treatment, auc_pl, auc_published, auc_pct_err,
         ratio_observed, ratio_published, ratio_pct_err) |>
  dplyr::rename(
    "Arm"                        = treatment,
    "AUC(0-24) solved"           = auc_pl,
    "Dose / 10.311"              = auc_published,
    "% error"                    = auc_pct_err,
    "Saliva:plasma solved"       = ratio_observed,
    "4.929 / 5.084"              = ratio_published,
    "% error "                   = ratio_pct_err
  ) |>
  knitr::kable(
    digits  = 4,
    caption = paste(
      "Steady-state closed-form gates against the Table 2 point estimates.",
      "Both identities are exact properties of the driven-effect-compartment",
      "structure; the residual is trapezoidal error on a 0.02 h grid."
    )
  )

# Deterministic solve against the published numbers. A tight bound is correct
# here and is what makes the gate catch a structural regression. Coupling the
# saliva state to central mass-balance-wise inflates total clearance ~134-fold
# (see Assumptions below), so this cannot pass under the wrong topology.
stopifnot(max(abs(cf_res$auc_pct_err)) < 0.5)
stopifnot(max(abs(cf_res$ratio_pct_err)) < 0.5)
```

The terminal half-life is a third deterministic identity, `ln2 * V / CL`,
and it is the sharpest single check on the two structural parameters because
it is dose-free.

```{r half-life-gate}
thalf_published <- log(2) * pub_vc / pub_cl
thalf_published

# Measured from the deterministic profile's own terminal slope, over the window
# where absorption has finished (Ka = 4.18 1/h, so absorption is complete within
# about 1 h of the 0.95 h lag).
thalf_solved <- cf |>
  filter(treatment == "750 mg q24h", tad >= 6) |>
  with(-log(2) / stats::coef(stats::lm(log(Cc) ~ tad))[["tad"]])
thalf_solved

stopifnot(abs(thalf_solved / thalf_published - 1) < 0.01)
```

The saliva:plasma exposure ratio is
`r round(cf_res$ratio_published[1], 4)`. The paper never states this number
directly, but it has an independent published anchor: the authors also fitted a
**scale-factor** saliva model to the same data, in which saliva concentration is
plasma concentration multiplied by a single estimated constant. Table S2 reports
that constant as **0.928**. Two structurally different models fitted to the same
paired samples should agree on the saliva:plasma ratio, and they do.

```{r ratio-vs-scale-factor}
scale_factor_S2 <- 0.928   # Table S2, competing scale-factor saliva model
c(separate_compartment = cf_res$ratio_published[1],
  scale_factor_model   = scale_factor_S2,
  pct_difference       = 100 * (cf_res$ratio_published[1] / scale_factor_S2 - 1))

# The two model forms are not required to agree exactly -- they were fitted
# separately and differ by dOFV = 4.95 -- but a large disagreement would mean one
# of the two saliva rate constants had been mis-transcribed. 10% is generous
# against the realised 4.5%, and swapping Kabs and Kel (0.9695 -> 1.0314) or
# dropping either would break it.
stopifnot(abs(cf_res$ratio_published[1] / scale_factor_S2 - 1) < 0.10)
```

Because the ratio does not depend on any subject-level parameter, it must also
be near-constant across the cohort. That is a genuine structural gate: if IIV
ever leaked onto either saliva rate constant, or if the saliva state picked up
its own volume, the spread would open up.

```{r ratio-cohort}
cohort_ratio <- sim |>
  group_by(id, treatment) |>
  summarise(ratio = trap(tad, Csaliva) / trap(tad, Cc), .groups = "drop")

range(cohort_ratio$ratio)
# On the coarse hourly grid the trapezoid biases the two AUCs slightly
# differently, and the bias varies a little with clearance, so this is a spread
# bound rather than an equality.
stopifnot(max(cohort_ratio$ratio) / min(cohort_ratio$ratio) < 1.10)
stopifnot(abs(median(cohort_ratio$ratio) / cf_res$ratio_published[1] - 1) < 0.05)
```

# Replicate published figures

```{r figure-3, fig.width = 7, fig.height = 4}
# Replicates Figure 3 of Nguyen 2026: steady-state concentration-time
# percentiles for plasma (left) and saliva (right). The paper's panels span the
# 0-5 h observed sampling window on a shared 0-22 mg/L axis with near-identical
# medians in the two matrices; the full 24 h interval is shown here since the
# model is used to extrapolate to AUC(0-24). The vertical span is matched to the
# paper's axis so the two are directly comparable by eye.
sim |>
  filter(treatment == "750 mg q24h") |>
  select(id, tad, Plasma = Cc, Saliva = Csaliva) |>
  tidyr::pivot_longer(c(Plasma, Saliva), names_to = "matrix",
                      values_to = "conc") |>
  group_by(matrix, tad) |>
  summarise(
    Q05 = quantile(conc, 0.05),
    Q50 = quantile(conc, 0.50),
    Q95 = quantile(conc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~matrix) +
  coord_cartesian(ylim = c(0, 22)) +
  labs(
    x = "Time after dose at steady state (h)",
    y = "Levofloxacin concentration (mg/L)",
    title = "Figure 3 - steady-state profiles by matrix, 750 mg once daily",
    caption = paste(
      "Median with 5th-95th percentile ribbon, 200 virtual subjects.",
      "Replicates Figure 3 of Nguyen 2026."
    )
  )
```

Figure 3's observed-median line is the only quantitative anchor the paper offers
for absolute concentrations, and it must be read off the plot rather than a
table. The three sampled timepoints give plasma medians of roughly 2.0, 5.3 and
5.0 mg/L at 0, 2 and 5 h post-dose, and saliva medians of roughly 2.0, 5.4 and
4.6 mg/L. These are **digitised**, so they are used only as a bracket: each
should fall between the deterministic 750 mg and 1000 mg typical-value curves,
since the study's own doses spanned exactly that range.

```{r figure-3-bracket}
# Digitised from Figure 3 of Nguyen 2026; see the caveat above. A 15% allowance
# covers the figure read, which is much wider than any plausible reading error
# on a line plotted against a 0-22 mg/L axis.
observed_median <- tibble::tribble(
  ~tad, ~matrix,   ~digitised,
     0, "Plasma",         2.0,
     2, "Plasma",         5.3,
     5, "Plasma",         5.0,
     0, "Saliva",         2.0,
     2, "Saliva",         5.4,
     5, "Saliva",         4.6
)

typical_at <- cf |>
  filter(tad %in% c(0, 2, 5)) |>
  select(treatment, tad, Plasma = Cc, Saliva = Csaliva) |>
  tidyr::pivot_longer(c(Plasma, Saliva), names_to = "matrix",
                      values_to = "typical") |>
  tidyr::pivot_wider(names_from = treatment, values_from = typical)

bracket_fig3 <- observed_median |>
  left_join(typical_at, by = c("tad", "matrix")) |>
  mutate(inside = digitised >= `750 mg q24h` * 0.85 &
                  digitised <= `1000 mg q24h` * 1.15)

bracket_fig3 |>
  dplyr::rename(
    "Time after dose (h)"      = tad,
    "Matrix"                   = matrix,
    "Figure 3 median (read)"   = digitised,
    "750 mg typical"           = `750 mg q24h`,
    "1000 mg typical"          = `1000 mg q24h`,
    "Inside bracket"           = inside
  ) |>
  knitr::kable(
    digits  = 2,
    caption = paste(
      "Digitised Figure 3 observed medians against the deterministic",
      "typical-value profiles at the study's two dose endpoints."
    )
  )

stopifnot(nrow(bracket_fig3) == 6L, all(bracket_fig3$inside))
```

Figure 4 and Tables 3 and S3 report maximum-a-posteriori re-estimation over 18
limited-sampling designs, each with 100 Monte Carlo replicates per subject.
Reproducing them requires refitting the model to simulated saliva data subject
by subject, which is outside what a validation vignette does; those results are
therefore not replicated here. What *is* checked is the model those simulations
were run from -- the structural identities above and the NCA table below.

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  # Only `!is.na(Cc)`: adding `tad > 0` or `Cc > 0` would drop the time-zero
  # trough record that PKNCA anchors AUC(0-tau) on.
  filter(!is.na(Cc)) |>
  select(id, time = tad, Cc, treatment) |>
  arrange(treatment, id, time)

dose_df <- events |>
  filter(evid == 1, time == t_last) |>
  mutate(time = 0) |>
  select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(
  sim_nca, Cc ~ time | treatment + id,
  concu = "mg/L", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  as.data.frame(dose_df), amt ~ time | treatment + id,
  doseu = "mg", route = "extravascular"
)

intervals <- data.frame(
  start     = 0,
  end       = tau,
  cmax      = TRUE,
  tmax      = TRUE,
  cmin      = TRUE,
  cav       = TRUE,
  auclast   = TRUE,
  half.life = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)
```

```{r pknca-saliva}
# Second PKNCA block for the saliva output (multi-output model).
sal_nca <- sim |>
  filter(!is.na(Csaliva)) |>
  select(id, time = tad, Cc = Csaliva, treatment) |>
  arrange(treatment, id, time)

nca_sal <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(sal_nca, Cc ~ time | treatment + id,
                     concu = "mg/L", timeu = "h"),
    dose_obj,
    intervals = intervals
  )
)
```

## Comparison against published values

Nguyen 2026 publishes **no** NCA table: Table 3 and Table S3 report mean
*differences* between saliva-predicted and reference plasma `AUC(0-24)` in
percent, not absolute exposures, and the abstract's only absolute claims are
the parameter estimates themselves. The comparator below is therefore
constructed from the Table 2 point estimates through the closed-form identities
of this model structure -- `AUC(0-24) = Dose / (CL/F)` and
`t(1/2) = ln2 * (V/F) / (CL/F)` -- with the published numbers hardcoded. This
is a weaker claim than a published-NCA comparison and is labelled as such, but
it is not circular: the reference row contains no model variable.

```{r nca-compare}
published_750 <- tibble::tribble(
  ~treatment,       ~auclast,                      ~half.life,
  "750 mg q24h",    750 / pub_cl,                  log(2) * pub_vc / pub_cl
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published_750,
  by            = "treatment",
  units         = c(auclast = "mg*h/L", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated 750 mg once-daily steady-state NCA against the closed-form",
    "values implied by Nguyen 2026 Table 2 (CL/F = 10.311 L/h,",
    "V/F = 278.88 L). * differs from reference by >20%."
  ),
  align = c("l", "l", "r", "r", "r")
)
attr(cmp, "footnote")
```

`ncaComparisonTable()` aggregates the per-subject PKNCA results with `median()`,
which matters here and is worth being explicit about. `AUC(0-24)` and half-life
are both proportional to `1/(CL/F)` and therefore log-normally distributed, so
their sample **mean** over a 200-subject cohort drifts several percent between
draws. Their sample **median** is both the robust statistic and, for this model,
the exactly predictable one: the median of `Dose/(CL/F)` over a log-normal
`CL/F` is `Dose/(CL/F)` at the typical value. Every assertion below is written
on medians or on deterministic typical-value quantities, never on a cohort mean.

```{r nca-assertions}
nca_tbl <- as.data.frame(nca_res$result) |>
  filter(treatment == "750 mg q24h") |>
  group_by(PPTESTCD) |>
  summarise(median = median(PPORRES), .groups = "drop")

# Guard against a silent zero-row lookup (a gate that cannot go red).
stopifnot(all(c("half.life", "tmax", "cmax", "auclast", "cav") %in%
                nca_tbl$PPTESTCD))
get_median <- function(code) {
  v <- nca_tbl$median[nca_tbl$PPTESTCD == code]
  if (length(v) != 1L) stop("no unique NCA row for '", code, "'")
  v
}

auc_pct   <- 100 * (get_median("auclast") / (750 / pub_cl) - 1)
thalf_pct <- 100 * (get_median("half.life") /
                      (log(2) * pub_vc / pub_cl) - 1)
c(auc_pct_diff = auc_pct, half_life_pct_diff = thalf_pct)

# Bounds on a cohort MEDIAN, not on an extreme. The sample median of a 41%-CV
# log-normal over 200 subjects carries roughly a 3.6% standard error, so 12% is
# about 3 SE of headroom -- while a mis-transcribed clearance or volume moves
# these by tens of percent and still goes red. `auclast` sits a little below
# Dose/(CL/F) because the hourly trapezoid under-reads the peak.
stopifnot(abs(auc_pct) < 12)
stopifnot(abs(thalf_pct) < 12)

# Dose proportionality is exact in this linear model and is a cheap check that
# the 1000 mg arm was simulated as intended.
auc_1000 <- as.data.frame(nca_res$result) |>
  filter(treatment == "1000 mg q24h", PPTESTCD == "auclast") |>
  pull(PPORRES) |>
  median()
stopifnot(abs(auc_1000 / get_median("auclast") / (1000 / 750) - 1) < 0.05)
```

The saliva NCA has no published comparator at all -- the paper reports saliva
only through model-predicted `AUC(0-24)` differences -- so it is reported for
completeness, with the saliva:plasma ratio already gated above:

```{r nca-saliva-summary}
as.data.frame(nca_sal$result) |>
  filter(PPTESTCD %in% c("auclast", "cmax", "tmax", "cmin", "cav")) |>
  group_by(treatment, PPTESTCD) |>
  summarise(
    median = median(PPORRES),
    q05    = quantile(PPORRES, 0.05),
    q95    = quantile(PPORRES, 0.95),
    .groups = "drop"
  ) |>
  dplyr::rename(
    "Arm"           = treatment,
    "NCA parameter" = PPTESTCD,
    "Median"        = median,
    "5th pctile"    = q05,
    "95th pctile"   = q95
  ) |>
  knitr::kable(
    digits = 2,
    caption = "Simulated saliva NCA over the steady-state dosing interval."
  )
```

# Assumptions and deviations

## Reconstruction of the saliva ODE (structural, load-bearing)

Nguyen 2026 publishes its control stream in Supplementary Table S4, but the
table is headed "**Example** of NONMEM model code" and is abridged: it declares
`$SUBROUTINE ADVAN13`, the general *nonlinear* solver, whose differential
equations must be written out in a `$DES` block -- and no `$DES` block is
printed. The two saliva rate constants therefore had to be assembled into an
ODE, and the `K23` / `K30` naming (borrowed from NONMEM's general *linear*
ADVAN5/ADVAN7 subroutines, where those names *do* imply a mass-balance matrix)
suggests a reading that is wrong.

The paper resolves it in prose. Methods 2.3 describes the saliva state as "a
saliva bio-compartment (i.e., a hypothetical effect compartment, **which does
not account for mass balance**)", and Figure 1's caption repeats that the dotted
saliva lines "represent a hypothetical effect compartment without a volume". The
central equation therefore carries no `-kin_saliva * central` loss term.

The arithmetic confirms it. Under the mass-balance reading, the saliva state
settles at `kin_saliva / kel_saliva = 0.9695` times the central amount, so loss
through saliva would run at `kel_saliva * 0.9695 = 4.929` per hour against a
plasma `kel` of `10.311 / 278.88 = 0.036973` per hour -- salivary elimination
would be **134 times** the plasma clearance. Steady-state `AUC(0-24)` on 750 mg
would collapse from 72.7 to roughly 0.5 `mg*h/L` and every simulated
concentration would sit two orders of magnitude below the 0.5-20 mg/L range
Figure 3 plots. The driven form is what the model file encodes.

**No printed parameter value was altered** to reach it -- all six structural
estimates are used exactly as published; only the unprinted `$DES` was
reconstructed. The same reading is already operator-ratified for this group's
companion linezolid model (`modellib("Nguyen_2026_linezolid")`, sidecar
`request-001` q2, 2026-09-02), which carries the identical Figure 1 language and
an identically abridged control stream.

## Saliva concentration scaling

Table S4's `$ERROR` block prints `CP = A(2)/V` for plasma but `CS = A(3)` for
saliva -- i.e. a saliva "concentration" equal to the saliva *amount*, an implied
1 L scale. Taken literally, the steady-state saliva:plasma exposure ratio
becomes `V * kin_saliva / kel_saliva = 278.88 * 0.9695 = 270`, so saliva would
run about 270 times plasma. That is contradicted twice over: Figure 3 plots both
matrices on one shared 0-22 mg/L axis with near-identical medians, and the
authors' own competing scale-factor model (Table S2) puts the ratio at 0.928.
Dividing by `vc` gives 0.9695, agreeing with that independent estimate to 4.5%.
The model encodes `Csaliva <- saliva / vc`, matching the ratified linezolid
precedent.

## Other assumptions

* **Dose levels.** The study's per-subject daily doses are not published, only
  the 750-1000 mg range and the 17.9 mg/kg/day median. The vignette simulates
  the two endpoints of that range as separate arms rather than guessing a mix.
* **No published NCA comparator.** Unlike most models in this library, the
  source paper reports no absolute Cmax / Tmax / `AUC(0-24)` values -- only
  percentage differences between sampling strategies. The comparison table is
  therefore built from the Table 2 point estimates through closed-form
  identities, with the published numbers hardcoded so a transcription error
  still goes red. The digitised Figure 3 bracket is the only absolute-scale
  check available and is treated accordingly.
* **Figure 3 medians are digitised.** The three observed medians per matrix were
  read off the plotted line, not a table. They are used as a bracket between the
  two dose arms with a 15% allowance, never as a target to match.
* **Weight is carried but unused.** No covariate was retained, so the drawn
  weights affect nothing. They are simulated to let the arms be reported in
  mg/kg alongside the paper's 15-20 mg/kg/day range.
* **No IIV on four parameters.** `Ka`, `Tlag`, `Kabs` and `Kel` carry no
  between-subject variability, because Table S4 fixes all four `$OMEGA` elements
  to zero and Results 3.2 explains that adding IIV on the two saliva rate
  constants "rendered the model unstable -- it terminated before minimisation
  and failed to report any significant digits". They are omitted from `ini()`
  rather than written as `~ fixed(0)`, which would make OMEGA singular and break
  rxode2's Cholesky sampler.
* **`Ka` and `Tlag` are fixed, not estimated.** Both are wrapped in `fixed()` so
  a user re-fitting the model inherits the authors' identifiability constraint
  rather than silently estimating a parameter the 0/2/5 h design cannot support.
  `Ka` was taken from published levofloxacin analyses rather than from this
  dataset at all; `Tlag` was estimated at 0.947 h in the plasma-only base model
  (Table S1) and held at the rounded 0.95 h once saliva was added.
* **`CL/F` and `V/F` are apparent.** Oral bioavailability was not identifiable,
  so both are relative to `F`. The Table S4 control stream carries no
  bioavailability THETA at all, so the model file adds no `lfdepot` term -- `F`
  is structurally 1 rather than an estimated-then-fixed anchor.
* **The volume is unusually large and the authors say so.** `V/F = 278.88 L` is
  about three times the roughly 90 L commonly reported for levofloxacin. The
  Discussion attributes this to the cohort's body composition and, primarily, to
  sampling that stopped at 5 h post-dose, which leaves the distribution and
  terminal phases poorly characterised and inflates the apparent volume.
  Consequently the model's terminal half-life of
  `r round(log(2) * 278.88 / 10.311, 1)` h is longer than the literature value
  for levofloxacin, and the model should be used for `AUC(0-24)` estimation --
  the purpose it was built for, and a quantity that depends only on `CL/F` --
  rather than for extrapolating late concentrations.
* **Screened covariates.** Body weight and fat-free mass were tested as
  allometric size descriptors and neither improved the fit; age, sex, renal and
  hepatic function markers were screened by stepwise covariate modelling and
  none were retained. All eight appear in the model file's
  `covariatesDataExcluded` metadata for provenance and are deliberately absent
  from `model()`. The cohort's renal function was uniformly normal, so the
  absence of a renal covariate is a property of this dataset and not evidence
  that renal function is unimportant for a predominantly renally cleared drug.
* **Bayesian limited-sampling results are not replicated.** Tables 3 and S3 and
  Figure 4 report maximum-a-posteriori re-estimation across 18 sampling designs;
  see the note under *Replicate published figures*.
* **No non-paper-derived parameter values.** Every `ini()` entry comes from
  Nguyen 2026 Table 2 or the Table S4 control stream. Nothing was supplied by
  correspondence or carried from an upstream model, and the only digitised
  quantities anywhere in this vignette are the six Figure 3 medians used in the
  bracket check, which feed no parameter.
