---
title: "Cemiplimab (Nguyen 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cemiplimab (Nguyen 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Nguyen_2022_cemiplimab"))$reference`
* Description: `r rxode2::rxode(readModelDb("Nguyen_2022_cemiplimab"))$description`
* Article: <https://doi.org/10.1002/psp4.12855> (open access; Appendix S1
  holds the NONMEM control stream of the final model and Tables S1-S11)

Nguyen 2022 updates the earlier cemiplimab population PK model (packaged
separately as `Yang_2021_cemiplimab`) on 1063 patients from four studies:
CSCC, BCC, NSCLC and the first-in-human study in advanced malignancies.
To stabilise the model the authors removed the IIV on Emax and T50, the
covariance between the CL/Q and V1/V2 random effects, and the proportional
residual-error component. The updated model was then validated externally
on 292 women with recurrent or metastatic cervical cancer (R/M CC) from
EMPOWER-Cervical 1, whose data were not used for estimation.

Structure: linear two-compartment model, 30-min IV infusion, with a
sigmoid-Emax change in clearance over time since the first dose:

\[
\mathrm{CL}_i(t) = \mathrm{CL}_{0,i}\,
  \exp\!\left(\frac{E_{\max,i}\, t^{\gamma}}{T_{50,i}^{\gamma} + t^{\gamma}}\right)
\]

with \(E_{\max} = -0.174\) (CL approaches \(e^{-0.174} = 0.840\) of
baseline), \(T_{50} = 73.7\) days and \(\gamma = 2.5\) (fixed). One random
effect is shared by CL and Q and one by V1 and V2.

The exposure-response part of the paper (Kaplan-Meier, Cox and logistic
regression analyses of OS, PFS, DOR, ORR and irAEs against first-dose
exposure) found no exposure-response relationship once baseline
albumin, body weight and ECOG status were adjusted for, and reports no
parametric exposure-response model, so only the PK model is packaged.

# Population

The model-building data set holds 17,312 post-dose concentrations from
1063 patients (Table S2): Study 1423 (first-in-human, advanced
malignancies, n = 398), Study 1540 (CSCC, n = 188), Study 1620 (BCC after
hedgehog-inhibitor therapy, n = 132) and Study 1624 (PD-L1 >= 50% NSCLC,
n = 345). Doses were 1, 3 or 10 mg/kg Q2W, 3 mg/kg Q3W, 200 mg Q2W or
350 mg Q3W, all as 30-min IV infusions (Table S1). Baseline demographics
(Table S4): median age 65 years (27-96), median weight 73.2 kg
(30.9-171.6), 70.6% male, 87.7% White, median albumin 38.6 g/L (20-93),
median ALT 19 IU/L (4-234).

The external-validation cohort (Table S5) was 292 women with R/M CC on
350 mg Q3W: median age 51 years, weight 61.9 kg (35.9-128.7), albumin
39.1 g/L (20-50.7), ALT 14 IU/L (4-101.5), 63.0% White and 29.1% Asian.

The same information is available programmatically via
`readModelDb("Nguyen_2022_cemiplimab")()$population`.

# Source trace

Every `ini()` value carries an in-file comment pointing to its source in
`inst/modeldb/specificDrugs/Nguyen_2022_cemiplimab.R`. Covariate equation
forms and reference values come from the Appendix S1 control stream; the
final estimates come from Table 1 (the `$THETA` block of Appendix S1
holds initial estimates only).

| Parameter / equation | Value | Source |
|---|---|---|
| `lcl` (TVCL0) | log(0.254 L/day) | Table 1 |
| `lq` (TVQ) | log(0.652 L/day) | Table 1 |
| `lvc` (TVV1) | log(3.35 L) | Table 1 |
| `lvp` (TVV2) | log(2.52 L) | Table 1 |
| `cl_time_max` (TVEmax) | -0.174 | Table 1 |
| `lcl_t50` (TVT50) | log(73.7 day) | Table 1 |
| `lcl_time_hill` (HILL) | log(2.50), fixed | Table 1, footnote c |
| `e_wt_cl_q` (CL_WGTBL) | 0.539 | Table 1 |
| `e_wt_vc_vp` (VssWGTBL) | 0.499 | Table 1 |
| `e_sexf_cl` (CL_SEX) | -0.137 | Table 1 |
| `e_sexf_vc` (V1_SEX) | -0.0801 | Table 1 |
| `e_alb_base_vc` (V1_ALBBL) | -0.217 | Table 1 |
| `e_alb_cl` (CL_ALB, time-varying) | -1.11 | Table 1 |
| `e_alt_cl` (CL_ALTBL) | -0.0729 | Table 1 |
| `e_tumtp_cscc_cl` (CL_CSCC) | -0.216 | Table 1 |
| `e_tumtp_bcc_cl` (CL_BCC) | -0.211 | Table 1 |
| `e_tumtp_bcc_cl_time_max` (Emax_BCC) | -0.872 | Table 1 |
| `e_tumtp_other_cl_t50` (T50_OTHER) | -0.491 | Table 1 |
| `etalcl` (ETA1, shared CL and Q) | 0.0892 | Table 1 |
| `etalvc` (ETA2, shared V1 and V2) | 0.0709 | Table 1 |
| `expSd` (RE) | 0.241 | Table 1, footnote d |
| Reference covariates: WT 75 kg, albumin 38.6 g/L, ALT 19 IU/L | -- | Appendix S1 `$PK` (`M_WGTBL`, `M_ALBBL`, `M_ALTBL`) |
| Power covariate forms; `(1 + theta)` forms for sex and tumor type | -- | Appendix S1 `$PK` |
| Q scaled by the CL weight exponent, V2 by the V1 weight exponent | -- | Appendix S1 `$PK` (`Q_COV = CLQ_WT`, `V2_COV = Vss_WT`) |
| Time-dependent CL, `exp(Emax * t^HILL / (T50^HILL + t^HILL))` | -- | Table 1 footnote; Appendix S1 `$PK` |
| Two-compartment ODEs (ADVAN3 TRANS4) | -- | Appendix S1 |
| Residual error additive on log concentration, `Y = LOG(CONC) + W*ERR(1)` | -- | Appendix S1 `$ERROR` |

# Virtual cohorts

The observed data are not public. Each study is represented by a virtual
cohort of 200 patients whose weight, albumin, ALT and sex follow the
per-study medians, ranges and percentages of Table S4 (Table S5 for
R/M CC). Weight and ALT are drawn log-normally and albumin normally, each
truncated to the published range; the joint covariate structure is not
known and is not reproduced. Time-varying albumin is held at its baseline
value (`ALB = ALB_BASE`). Every patient receives 350 mg Q3W for 17 doses
(the regimen of Table S7).

```{r cohort}
set.seed(2022)
n_per <- 200
tau <- 21
n_dose <- 17
dose_times <- seq(0, by = tau, length.out = n_dose)

draw_trunc <- function(n, rfun, lo, hi) pmin(pmax(rfun(n), lo), hi)

# Per-study covariate summaries from Table S4 (Table S5 for R/M CC).
cohort_spec <- tibble::tribble(
  ~cohort,                   ~wt_med, ~wt_lo, ~wt_hi, ~alb_med, ~alb_lo, ~alb_hi, ~alt_med, ~alt_lo, ~alt_hi, ~male_pct, ~cscc, ~bcc, ~other,
  "R/M CC (Study 1676)",       61.9,   35.9,  128.7,     39.1,      20,    50.7,       14,       4,   101.5,         0,     0,    0,      0,
  "CSCC (Study 1540)",         80.7,   31.0,  171.6,     40.0,      25,    48.0,       17,       6,   234.0,      83.0,     1,    0,      0,
  "BCC (Study 1620)",          73.2,   44.6,  134.8,     41.0,      28,    93.0,       16,       5,    69.0,      67.4,     0,    1,      0,
  "NSCLC (Study 1624)",        69.6,   37.6,  138.0,     38.9,      20,    53.5,       17,       4,   133.0,      87.5,     0,    0,      0,
  "All solid tumors (Study 1423)", 75.3, 30.9, 156.0,    37.0,      22,    48.0,       22,       5,   196.0,      51.3,     0,    0,      1
)

make_cohort <- function(spec, id_offset) {
  n <- n_per
  alb <- draw_trunc(n, function(k) rnorm(k, spec$alb_med, 5), spec$alb_lo, spec$alb_hi)
  tibble(
    id = id_offset + seq_len(n),
    cohort = spec$cohort,
    WT = draw_trunc(n, function(k) rlnorm(k, log(spec$wt_med), 0.2), spec$wt_lo, spec$wt_hi),
    ALB_BASE = alb,
    ALB = alb,
    ALT = draw_trunc(n, function(k) rlnorm(k, log(spec$alt_med), 0.5), spec$alt_lo, spec$alt_hi),
    SEXF = rbinom(n, 1, 1 - spec$male_pct / 100),
    TUMTP_CSCC = spec$cscc,
    TUMTP_BCC = spec$bcc,
    TUMTP_OTHER = spec$other
  )
}

subjects <- dplyr::bind_rows(lapply(seq_len(nrow(cohort_spec)), function(i) {
  make_cohort(cohort_spec[i, ], id_offset = (i - 1L) * n_per)
}))
stopifnot(!anyDuplicated(subjects$id))

# Observation grid: dense over the first and the last (steady-state) dosing
# interval, including the end of each 30-min infusion and the trough.
eoi <- 0.5 / 24
ss_start <- dose_times[n_dose]
obs_times <- sort(unique(c(
  0, eoi, seq(1, tau, by = 1),
  seq(tau, ss_start, by = 3),
  ss_start + c(eoi, seq(1, tau, by = 1))
)))

doses <- subjects |>
  tidyr::crossing(time = dose_times) |>
  dplyr::mutate(evid = 1L, amt = 350, dur = eoi, cmt = "central")
obs <- subjects |>
  tidyr::crossing(time = obs_times) |>
  dplyr::mutate(evid = 0L, amt = 0, dur = 0, cmt = "central")
events <- dplyr::bind_rows(doses, obs) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  dplyr::select(id, time, evid, amt, dur, cmt, dplyr::everything()) |>
  as.data.frame()
```

# Simulation

Table S7 reports *population-predicted* exposures, i.e. each patient's
typical-value profile at their own covariates with the random effects set
to zero. The comparison below therefore uses `zeroRe()`; a stochastic
simulation is used afterwards for the variability-dependent quantities.

```{r simulate}
mod <- readModelDb("Nguyen_2022_cemiplimab")
mod_typ <- rxode2::zeroRe(mod)
sim_typ <- rxode2::rxSolve(
  mod_typ, events = events, keep = "cohort", returnType = "data.frame"
)
stopifnot(!anyNA(sim_typ$Cc))
```

## Typical concentration-time profiles by tumor type

```{r profile-plot, fig.width = 7, fig.height = 4.5}
sim_typ |>
  dplyr::group_by(cohort, time) |>
  dplyr::summarise(Cc = mean(Cc), .groups = "drop") |>
  ggplot(aes(time, Cc, colour = cohort)) +
  geom_line() +
  labs(
    x = "Time since first dose (day)",
    y = "Cemiplimab concentration (mg/L)",
    colour = NULL,
    title = "Mean population-predicted profiles, 350 mg Q3W",
    caption = "Grid is coarse (every 3 days) between the first and the last dosing interval."
  ) +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(ncol = 2))
```

## Time-dependent clearance

Typical CL(t) / CL(0) for the reference patient and for the two tumor
types that carry an effect on the time course: BCC (Emax reduced by
87.2%) and the other tumor types (T50 reduced by 49.1%).

```{r cl-time, fig.width = 6, fig.height = 4}
cl_ratio <- sim_typ |>
  dplyr::filter(cohort %in% c("NSCLC (Study 1624)", "BCC (Study 1620)", "All solid tumors (Study 1423)")) |>
  dplyr::group_by(cohort, time) |>
  dplyr::summarise(ratio = mean(cl / cl_base), .groups = "drop")

ggplot(cl_ratio, aes(time, ratio, colour = cohort)) +
  geom_line() +
  geom_hline(yintercept = exp(-0.174), linetype = "dashed", colour = "grey40") +
  labs(x = "Time since first dose (day)", y = "CL(t) / CL(0)", colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")

# The asymptote is exp(Emax): 0.840 for the reference and exp(-0.174 *
# 0.128) = 0.978 for BCC. Deterministic (typical values), so the bounds are tight.
ratio_end <- cl_ratio |> dplyr::filter(time == max(time))
stopifnot(
  abs(ratio_end$ratio[ratio_end$cohort == "BCC (Study 1620)"] - exp(-0.174 * 0.128)) < 0.002,
  abs(ratio_end$ratio[ratio_end$cohort == "NSCLC (Study 1624)"] - exp(-0.174)) < 0.01
)
```

Nguyen 2022 reports a mean 10.8% decrease in CL from the first dose to
steady state, generally similar across tumor types except BCC, for which
the change was minimal (Table 2: +2.7%). The typical-value curves show
the same pattern: about -16% for the reference and other tumor types
by the end of the first year, and about -2% for BCC. The post hoc
means of Table 2 are smaller because they average over individual
albumin trajectories and an unstated steady-state time.

# PKNCA: population-predicted exposure (Table S7)

```{r pknca}
conc <- sim_typ |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, cohort)
dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, cohort)

conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ time | cohort + id, concu = "mg/L", timeu = "day")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | cohort + id, doseu = "mg")
intervals <- data.frame(
  start = c(0, ss_start),
  end = c(tau, ss_start + tau),
  cmax = TRUE,
  clast.obs = TRUE,
  auclast = TRUE
)
nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

The trough is the concentration at the end of each 21-day interval
(`clast.obs`); `cmin` would return the pre-dose zero of the first
interval. Table S7 reports means across patients, so the simulated
per-patient values are averaged (not pooled by median) before the
comparison.

```{r nca-compare}
sim_mean <- as.data.frame(nca$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "clast.obs", "auclast")) |>
  dplyr::mutate(interval = ifelse(start == 0, "First dose", "Steady state")) |>
  dplyr::group_by(cohort, interval, PPTESTCD) |>
  dplyr::summarise(PPORRES = mean(PPORRES), .groups = "drop")

# Table S7, population-predicted means for 350 mg Q3W.
ref <- tibble::tribble(
  ~cohort,                          ~interval,      ~auclast, ~cmax, ~clast.obs,
  "R/M CC (Study 1676)",            "First dose",        993,   125,  27.3,
  "CSCC (Study 1540)",              "First dose",        949,   113,  27.0,
  "BCC (Study 1620)",               "First dose",        967,   109,  29.3,
  "NSCLC (Study 1624)",             "First dose",        892,   117,  22.8,
  "All solid tumors (Study 1423)",  "First dose",        865,   108,  23.1,
  "R/M CC (Study 1676)",            "Steady state",     1949,   185,  60.1,
  "CSCC (Study 1540)",              "Steady state",     2111,   182,  69.5,
  "BCC (Study 1620)",               "Steady state",     2027,   176,  66.4,
  "NSCLC (Study 1624)",             "Steady state",     1777,   171,  53.8,
  "All solid tumors (Study 1423)",  "Steady state",     1770,   164,  55.2
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_mean,
  reference = ref,
  by = c("cohort", "interval"),
  units = c(cmax = "mg/L", clast.obs = "mg/L", auclast = "day*mg/L")
)
knitr::kable(cmp, caption = "Simulated vs. Table S7 mean population-predicted exposure, 350 mg Q3W (AUC over one 21-day interval; Clast is the trough at day 21 of the interval).")
if (!is.null(attr(cmp, "footnote"))) cat(attr(cmp, "footnote"))

pct <- sim_mean |>
  dplyr::inner_join(
    tidyr::pivot_longer(ref, c(auclast, cmax, clast.obs), names_to = "PPTESTCD", values_to = "ref"),
    by = c("cohort", "interval", "PPTESTCD")
  ) |>
  dplyr::mutate(pct_diff = 100 * (PPORRES / ref - 1))
stopifnot(nrow(pct) == 30L)
# The covariate cohorts are random draws from marginal summaries, so assert
# on the centre and a robust envelope (CLAUDE.md, "Vignette assertions over a
# simulated cohort"). A mis-transcribed CL, V1 or unit moves every row by
# tens of percent.
stopifnot(
  abs(median(pct$pct_diff)) < 7,
  quantile(abs(pct$pct_diff), 0.9) < 15
)
```

The steady-state interval used here is the 17th dose (days 336-357), by
which CL has essentially reached its asymptote; the paper does not state
which interval its "steady state" refers to. The largest residual gaps are
expected for Study 1423, whose patients are represented here entirely as
"other tumor types", whereas the real study also enrolled NSCLC and CSCC
patients (it had CSCC expansion cohorts). First-dose exposures agree
within about 10% for every cohort, and steady-state exposures within
about 5% for the four model-building studies. For R/M CC the simulated
steady-state exposures sit roughly 5-11% above Table S7 while the
first-dose values match. One possible cause is the time-varying
albumin on CL, which is held at baseline here; another is the
covariate joint distribution of the real cervical-cancer cohort, which
the independent marginal draws do not reproduce.

# Post hoc parameters with IIV (Table 2)

A stochastic simulation of the four model-building cohorts reproduces the
Table 2 post hoc summaries: CL at first dose, Vss and the first-dose
terminal half-life computed from the two-compartment micro-constants.

```{r posthoc}
first <- subjects |>
  dplyr::filter(cohort != "R/M CC (Study 1676)") |>
  dplyr::mutate(time = 0, evid = 0L, amt = 0, dur = 0, cmt = "central") |>
  dplyr::select(id, time, evid, amt, dur, cmt, dplyr::everything()) |>
  as.data.frame()
sim_iiv <- rxode2::rxSolve(mod, events = first, keep = "cohort", returnType = "data.frame")
ph <- sim_iiv |>
  dplyr::mutate(
    vss = vc + vp,
    beta = 0.5 * ((kel + k12 + k21) - sqrt((kel + k12 + k21)^2 - 4 * kel * k21)),
    thalf = log(2) / beta
  )
ph_tab <- dplyr::bind_rows(
  ph |> dplyr::mutate(cohort = "All patients"),
  ph
) |>
  dplyr::group_by(cohort) |>
  dplyr::summarise(
    cl_mean = mean(cl), cl_cv = 100 * sd(cl) / mean(cl),
    vss_mean = mean(vss), thalf_mean = mean(thalf),
    .groups = "drop"
  )
ph_ref <- tibble::tribble(
  ~cohort,                          ~cl_ref, ~vss_ref, ~thalf_ref,
  "All patients",                     0.246,     5.89,       19.7,
  "CSCC (Study 1540)",                0.214,     5.86,       21.4,
  "BCC (Study 1620)",                 0.192,     6.07,       24.9,
  "NSCLC (Study 1624)",               0.265,     5.61,       17.5,
  "All solid tumors (Study 1423)",    0.262,     6.10,       19.2
)
ph_cmp <- dplyr::inner_join(ph_ref, ph_tab, by = "cohort")
ph_cmp |>
  dplyr::transmute(
    Cohort = cohort,
    `CL first dose, Table 2 (L/day)` = cl_ref,
    `CL first dose, simulated (L/day)` = signif(cl_mean, 3),
    `CL CV%, simulated` = round(cl_cv, 1),
    `Vss, Table 2 (L)` = vss_ref,
    `Vss, simulated (L)` = signif(vss_mean, 3),
    `t1/2 first dose, Table 2 (day)` = thalf_ref,
    `t1/2 first dose, simulated (day)` = signif(thalf_mean, 3)
  ) |>
  knitr::kable(caption = "Table 2 post hoc means vs. simulated means (random effects included).")

ph_pct <- with(ph_cmp, c(
  100 * (cl_mean / cl_ref - 1),
  100 * (vss_mean / vss_ref - 1),
  100 * (thalf_mean / thalf_ref - 1)
))
stopifnot(length(ph_pct) == 15L, abs(median(ph_pct)) < 8, quantile(abs(ph_pct), 0.9) < 20)
```

The overall CV of CL at first dose in Table 2 is 40.9%; the simulated CV
combines the 30.5% CV of the shared CL/Q random effect with the covariate
spread of the virtual cohorts.

# External validation cohort: R/M CC

Figure 3 of Nguyen 2022 is an external posterior predictive check of the
model against 2030 concentrations from 292 R/M CC patients on 350 mg Q3W.
The band below is the corresponding model prediction for the virtual R/M
CC cohort, with IIV and residual error.

```{r rmcc-vpc, fig.width = 7, fig.height = 4.5}
ev_cc <- events |> dplyr::filter(cohort == "R/M CC (Study 1676)", time <= 126)
sim_cc <- rxode2::rxSolve(mod, events = ev_cc, keep = "cohort", returnType = "data.frame")
sim_cc |>
  dplyr::filter(time > 0) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    med = median(sim), lo = quantile(sim, 0.05), hi = quantile(sim, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, med)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.25) +
  geom_line() +
  scale_y_log10() +
  labs(
    x = "Time since first dose (day)",
    y = "Cemiplimab concentration (mg/L)",
    title = "R/M CC, 350 mg Q3W: median and 90% prediction interval",
    caption = "Compare with Figure 3 of Nguyen 2022. N = 200 virtual patients."
  ) +
  theme_bw()
```

# Assumptions and deviations

- **Final estimates vs. initial estimates.** The `$THETA` block in
  Appendix S1 lists initial estimates (e.g. TVCL 0.263, TVV2 1.91); the
  packaged values are the final estimates of Table 1. The control stream
  supplies the equation forms and the reference covariate values only.
- **Covariate parameters fixed to zero in the control stream** (age,
  AST, bilirubin, creatinine clearance, serum creatinine, ECOG, ethnicity,
  race, CSCC and other-tumor effects on Emax, albumin effects on Emax and
  T50) are absent from Table 1 and are not encoded.
- **Signed Emax.** `cl_time_max` is negative (CL decreases over time), so
  it stays on the linear scale; the convention checker's log-transform
  warning for this parameter is expected.
- **Tumor-type coding.** The control stream codes CSCC as `TUMTYPN = 1`,
  BCC as 2 and "OTHERS" as 5, with no effect for the remaining codes,
  which include NSCLC. The reference category is therefore NSCLC, and the
  "other" pool, whose composition the paper does not list, shares the
  reference CL and Emax but has a shorter T50. The paper does not state
  which code the R/M CC patients carried in the external validation. The
  vignette treats them as the reference category; coding them as "other"
  only shortens T50, which changes steady-state CL by about 4%.
- **Study 1423 cohort.** All virtual Study 1423 patients are coded as
  "other tumor types". The real study also enrolled CSCC and NSCLC
  patients, in proportions the paper does not report.
- **Albumin.** The model uses time-varying albumin on CL and baseline
  albumin on V1 as separate columns (`ALB`, `ALB_BASE`). Albumin
  trajectories are not reported, so the simulations hold `ALB` at the
  baseline value.
- **Table S7 AUC units.** Table S7 labels AUC3wks as mg/L; the
  values are areas over a 21-day interval and are compared here in
  `day*mg/L`.
- **Weight reference.** The control stream normalises weight to 75 kg
  (`M_WGTBL = 75`) although the cohort median was 73.2 kg; the packaged
  model uses 75 kg as published.
- **Time origin.** Time-dependent CL is driven by time since the first
  dose (`ABS(TIME)` in the control stream). The packaged model uses
  rxode2's `t`, so the first dose must be given at time 0.
- **Virtual covariates.** Weight, albumin and ALT are drawn independently
  from marginal distributions anchored at the Table S4 / S5 medians and
  ranges; their spreads (log-SD 0.2 for weight, SD 5 g/L for albumin,
  log-SD 0.5 for ALT) were chosen by the maintainers, not reported.
