---
title: "Sufentanil (Nie 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sufentanil (Nie 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Nie_2025_sufentanil"))$reference`
* Description: `r rxode2::rxode(readModelDb("Nie_2025_sufentanil"))$description`
* Article: <https://doi.org/10.2147/DDDT.S500189>

This is the first published population PK model of epidural sufentanil in
labouring women. Its purpose is not maternal exposure -- which is very low --
but the quantification of *placental transfer*: the model carries an explicit
umbilical cord compartment so that fetal-side exposure can be predicted at, and
after, delivery.

# Population

Forty-one primiparous women contributed to the final population PK analysis, a
subset of the 90 parturients recruited between November 2018 and February 2019
across six Chinese centres (Nie 2025 Results, "Patient Characteristics"). They
were 27.2 +/- 2.9 years old and weighed 70.2 +/- 7.5 kg, with ASA physical
status II-III, BMI 18.5-30 kg/m^2, a singleton pregnancy and cervical dilation
of 2 <= phi < 6 cm at the time of analgesia.

The analysis dataset held 79 maternal central-compartment concentrations and 37
umbilical cord concentrations. Maternal venous samples were scheduled at 0.5 h
and 1 h after the epidural loading dose, at delivery, and 2 h after delivery;
a single umbilical venous sample was drawn immediately after cord clamping. Not
all four maternal samples were mandatory, so the design is sparse and
unbalanced. Sufentanil was assayed by LC-MS/MS with a lower limit of
quantification of 1 pg/mL.

No covariate model was reported. Nie 2025 Table 1 contains structural,
inter-individual-variability and residual rows only, and the paper describes no
covariate screening step, so the packaged model carries no `covariateData`.

The same information is available programmatically via
`readModelDb("Nie_2025_sufentanil")()$population`.

```{r population}
pop <- rxode2::rxode(readModelDb("Nie_2025_sufentanil"))$population
str(pop[c("species", "n_subjects", "n_studies", "age_mean", "weight_mean")])
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Nie_2025_sufentanil.R` carries an in-file comment
naming its origin. They are collected here for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL) | 176 L/h (RSE 9.25%) | Table 1, "Structural parameters" |
| `lvc` (V1) | 519 L (RSE 6.45%) | Table 1, "Structural parameters" |
| `lqcord` (CL2) | 0.0134 L/h (RSE 43.2%) | Table 1, "Structural parameters" |
| `lvcord` (V2) | 0.187 L (RSE 39.9%) | Table 1, "Structural parameters" |
| `etalcl` | 0.255046 = log(0.539^2 + 1) | Table 1, "IIV of CL" = 53.9 %CV (RSE 12%) |
| `etalvc` | 0.046429 = log(0.218^2 + 1) | Table 1, "IIV of V1" = 21.8 %CV (RSE 26.9%) |
| `propSd` | 0.206 | Table 1, "eps prop1 (%)" = 20.6% (RSE 13.4%) |
| `propSd_Ccord` | 0.355 | Table 1, "eps prop2 (%)" = 35.5% (RSE 12%) |
| `kel <- cl / vc` | n/a | Methods, "Sufentanil Structural Pharmacokinetic Model": K10 = CL/V1 |
| `k12 <- qcord / vc` | n/a | Methods, same section: K12 = CL2/V1 |
| `k21 <- qcord / vcord` | n/a | Methods, same section: K21 = CL2/V2 |
| `d/dt(central)`, `d/dt(cord)` | n/a | Methods, same section (A1, A2 mass balance); Figure 1 schematic |
| Dose enters `central` | n/a | Methods, same section: the epidural compartment "merged with the central compartment" |
| Exponential IIV, log-normal | n/a | Methods, "Population PK Model Development" |
| Proportional-only residual | n/a | Methods, same section: "eps_add was removed due to its lesser significance" |

The IIV magnitudes in Table 1 are reported as %CV, so the internal variances are
`omega^2 = log(CV^2 + 1)`. That the conversion round-trips is asserted below.

```{r iiv-roundtrip}
ui <- rxode2::rxode(readModelDb("Nie_2025_sufentanil"))
omegas <- c(etalcl = 0.255046, etalvc = 0.046429)
roundtrip <- sqrt(exp(omegas) - 1) * 100
stopifnot(abs(roundtrip - c(53.9, 21.8)) < 0.05)
round(roundtrip, 2)
```

# Dosing regimen

The protocol (Nie 2025 Methods, "Anesthesia Management") is a 5 mL epidural
injection of the study mixture, a 10 mL bolus five minutes later, and
patient-controlled epidural analgesia from 30 min with a 6 mL/h continuous
infusion, a 6 mL demand bolus and a 15 min lockout, maintained until delivery.

The **sufentanil concentration of that mixture is reported inconsistently**: the
Abstract states 0.3 ug/mL, while the Methods state 3 ug/mL. The two differ
tenfold, and the model's own parameters settle it. At 6 mL/h the steady-state
maternal concentration is the infusion rate divided by CL:

```{r dose-falsifier}
CL <- 176; V1 <- 519; CL2 <- 0.0134; V2 <- 0.187
falsifier <- tibble::tibble(
  mixture_ug_per_mL = c(0.3, 3),
  css_pg_per_mL     = mixture_ug_per_mL * 6 * 1000 / CL,
  loading_c0_pg_per_mL = mixture_ug_per_mL * 15 * 1000 / V1
)
falsifier |>
  dplyr::rename(
    "Mixture (ug/mL)"                  = mixture_ug_per_mL,
    "Maternal Css at 6 mL/h (pg/mL)"   = css_pg_per_mL,
    "C0 after the 15 mL loading (pg/mL)" = loading_c0_pg_per_mL
  ) |>
  knitr::kable(digits = 1, caption = "Which mixture concentration is consistent with Table 1?")
```

Observed maternal concentrations in Nie 2025 Figure 2 (left panel) span roughly
3-30 pg/mL, with the 0.5-1 h loading cluster around 5-15 pg/mL. Only the
0.3 ug/mL mixture reproduces that; 3 ug/mL predicts a plateau of ~102 pg/mL,
which is above every observed point in the paper. **This vignette therefore uses
0.3 ug/mL** and treats the Methods figure as a typographical error. 0.3 ug/mL is
also the clinically conventional strength for epidural labour analgesia.

```{r regimen}
MIX_UG_PER_ML <- 0.3       # sufentanil concentration of the epidural mixture
LOAD1  <- 5  * MIX_UG_PER_ML * 1000   # ng, 5 mL initial injection at t = 0
LOAD2  <- 10 * MIX_UG_PER_ML * 1000   # ng, 10 mL bolus at t = 5 min
T_LOAD2 <- 5 / 60                     # h
T_PCEA  <- 0.5                        # h, PCEA started 30 min after loading
INF_RATE <- 6 * MIX_UG_PER_ML * 1000  # ng/h, 6 mL/h continuous infusion
LLOQ <- 1                             # pg/mL (Methods, assay section)
c(LOAD1 = LOAD1, LOAD2 = LOAD2, INF_RATE = INF_RATE)
```

Amounts are in ng and volumes in L, so `amount / volume` is ng/L, which is
identical to the pg/mL used throughout the paper. No conversion constant is
needed anywhere in the model.

# Virtual cohort

Original observed data are not available. Two virtual cohorts are used, each
well under the 200-per-arm cap.

* **Typical-value profile** -- one subject with the random effects zeroed, on a
  dense time grid, to reproduce the shape of the individual fits in Figure 4.
* **Figure 2 cohort** -- 200 women sampled exactly as the trial sampled them
  (maternal at 0.5 h, 1 h, delivery and delivery + 2 h; a single cord sample at
  delivery), with each woman's delivery time drawn over the 2.5-10.5 h range
  spanned by the observed sampling times in Figure 2.

```{r cohort}
set.seed(20250213)
N_COHORT <- 200

subj <- tibble::tibble(
  id       = seq_len(N_COHORT),
  delivery = runif(N_COHORT, 2.5, 10.5)   # h, epidural infusion stops at delivery
)

# Dosing is identical for everyone except the infusion duration, which runs from
# T_PCEA until that woman's delivery. `rate` + `amt` encodes the zero-order
# infusion; `delivery` rides along so rxSolve(keep = ) can return it.
dosing <- dplyr::bind_rows(
  subj |> dplyr::transmute(id, delivery, time = 0, amt = LOAD1, rate = 0,
                           evid = 1L, cmt = "central"),
  subj |> dplyr::transmute(id, delivery, time = T_LOAD2, amt = LOAD2, rate = 0,
                           evid = 1L, cmt = "central"),
  subj |> dplyr::transmute(id, delivery, time = T_PCEA,
                           amt = INF_RATE * (delivery - T_PCEA), rate = INF_RATE,
                           evid = 1L, cmt = "central")
)

# Observation rows use the DECLARED endpoint names (`Cc`, `Ccord`). This model
# declares two endpoints, so rxode2 allocates a slot for each and the event
# table must say which endpoint a row belongs to; using an ODE state name here
# instead raises "'dvid'->'cmt' on observation record".
obs <- dplyr::bind_rows(
  subj |> tidyr::crossing(t_fixed = c(0.5, 1)) |>
    dplyr::transmute(id, delivery, time = t_fixed, amt = NA_real_, rate = 0,
                     evid = 0L, cmt = "Cc"),
  subj |> dplyr::transmute(id, delivery, time = delivery, amt = NA_real_, rate = 0,
                           evid = 0L, cmt = "Cc"),
  subj |> dplyr::transmute(id, delivery, time = delivery + 2, amt = NA_real_, rate = 0,
                           evid = 0L, cmt = "Cc"),
  subj |> dplyr::transmute(id, delivery, time = delivery, amt = NA_real_, rate = 0,
                           evid = 0L, cmt = "Ccord")
)

events <- dplyr::bind_rows(dosing, obs) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid", "cmt")])))
nrow(events)
```

# Simulation

`useLinCmt = FALSE` is required: rxode2's automatic ODE-to-`linCmt()` conversion
corrupts the endpoint-to-compartment mapping for multi-output models of this
shape.

```{r simulate}
mod <- readModelDb("Nie_2025_sufentanil")

sim_cohort <- rxode2::rxSolve(
  mod, events = events, keep = "delivery", useLinCmt = FALSE
) |>
  as.data.frame()

# CMT 3 is the maternal endpoint (Cc), CMT 4 the umbilical cord endpoint
# (Ccord). `Cc` / `Ccord` are the individual predictions; `sim` carries the
# proportional residual error and is the simulated *observation*.
sim_cohort <- sim_cohort |>
  dplyr::mutate(
    endpoint = dplyr::if_else(CMT == 3, "Maternal plasma", "Umbilical cord"),
    ipred    = dplyr::if_else(CMT == 3, Cc, Ccord)
  )

stopifnot(dplyr::n_distinct(sim_cohort$id) == N_COHORT)
```

The typical-value profile uses a fixed 8 h infusion -- a representative labour
duration inside the observed range -- on a dense grid.

```{r simulate-typical}
grid_dense <- c(seq(0, 12, by = 0.05), seq(12.25, 48, by = 0.25))
INF_DUR <- 8

ev_typ <- rxode2::et(amt = LOAD1, time = 0, cmt = "central") |>
  rxode2::et(amt = LOAD2, time = T_LOAD2, cmt = "central") |>
  rxode2::et(amt = INF_RATE * INF_DUR, time = T_PCEA, dur = INF_DUR, cmt = "central") |>
  rxode2::et(grid_dense, cmt = "Cc") |>
  rxode2::et(grid_dense, cmt = "Ccord")

sim_typ <- rxode2::rxSolve(rxode2::zeroRe(mod), events = ev_typ, useLinCmt = FALSE) |>
  as.data.frame() |>
  dplyr::filter(CMT == 3)   # both Cc and Ccord are returned on every row
```

# Replicate published figures

## Figure 4 -- individual fits: maternal falls fast, cord does not

Figure 4 of Nie 2025 shows, for each woman, a maternal profile (solid black)
that plateaus during the epidural infusion and then falls steeply once the
infusion stops at delivery, alongside a cord profile (dashed red) that rises
slowly throughout labour and then stays almost flat. The typical-value
prediction reproduces both.

```{r figure-4}
sim_typ |>
  dplyr::select(time, `Maternal plasma` = Cc, `Umbilical cord` = Ccord) |>
  tidyr::pivot_longer(-time, names_to = "endpoint", values_to = "conc") |>
  dplyr::filter(conc > 0) |>
  ggplot(aes(time, conc, colour = endpoint, linetype = endpoint)) +
  geom_vline(xintercept = T_PCEA + INF_DUR, linetype = "dotted") +
  geom_line(linewidth = 0.9) +
  annotate("text", x = T_PCEA + INF_DUR + 0.4, y = 0.02,
           label = "infusion stops (delivery)", hjust = 0, size = 3) +
  scale_y_log10() +
  scale_colour_manual(values = c("Maternal plasma" = "black", "Umbilical cord" = "red")) +
  labs(x = "Time (h)", y = "Sufentanil concentration (pg/mL)",
       colour = NULL, linetype = NULL,
       title = "Typical-value maternal and umbilical cord profiles",
       caption = "Replicates the shape of Figure 4 of Nie 2025 (8 h infusion).")
```

The paper's qualitative claims are asserted numerically below.

```{r figure-4-assertions}
at_time <- function(tt, col) sim_typ[[col]][which.min(abs(sim_typ$time - tt))]
t_stop <- T_PCEA + INF_DUR

cord_peak_time <- sim_typ$time[which.max(sim_typ$Ccord)]
fig4 <- tibble::tibble(
  claim = c(
    "Maternal plateau during infusion (pg/mL)",
    "Cord concentration at delivery (pg/mL)",
    "Cord peaks AFTER the infusion stops (h past stop)",
    "Maternal concentration 16 h after stop (pg/mL)",
    "Cord concentration 16 h after stop, as % of its peak"
  ),
  value = c(
    at_time(t_stop, "Cc"),
    at_time(t_stop, "Ccord"),
    cord_peak_time - t_stop,
    at_time(t_stop + 16, "Cc"),
    100 * at_time(t_stop + 16, "Ccord") / max(sim_typ$Ccord)
  )
)

# The paper's central finding: the cord compartment keeps filling after the
# infusion stops, and is still near its peak when maternal drug has gone.
stopifnot(cord_peak_time > t_stop)
stopifnot(at_time(t_stop + 16, "Cc") < 0.1)
stopifnot(100 * at_time(t_stop + 16, "Ccord") / max(sim_typ$Ccord) > 40)

fig4 |>
  dplyr::rename("Claim from Nie 2025" = claim, "Simulated" = value) |>
  knitr::kable(digits = 3, caption = "Numerical restatement of the Figure 4 narrative.")
```

## Figure 2 -- observed concentration-time scatter

Figure 2 plots every observed concentration against time, maternal on the left
and umbilical cord on the right. Simulating the trial's own sparse sampling
design reproduces both clouds. Simulated observations below the 1 pg/mL assay
limit are dropped, as they would have been in the real dataset.

```{r figure-2}
sim_obs <- sim_cohort |>
  dplyr::filter(sim >= LLOQ)

ggplot(sim_obs, aes(time, sim)) +
  geom_point(shape = 1, alpha = 0.5) +
  facet_wrap(~endpoint) +
  scale_y_log10(limits = c(0.2, 60)) +
  labs(x = "Time (h)", y = "Sufentanil concentration (pg/mL)",
       title = "Simulated concentrations under the trial's sampling design",
       caption = "Replicates Figure 2 of Nie 2025 (left: maternal; right: umbilical cord).")
```

```{r figure-2-ranges}
# Ranges read off the log axis of Nie 2025 Figure 2 by the operator (graphical
# digitisation, not a printed table). Used only as a validation envelope -- no
# model parameter is derived from them.
observed_envelope <- tibble::tribble(
  ~endpoint,          ~obs_low, ~obs_high,
  "Maternal plasma",  3,        30,
  "Umbilical cord",   0.6,      10
)

sim_obs |>
  dplyr::group_by(endpoint) |>
  dplyr::summarise(
    sim_p05 = quantile(sim, 0.05),
    sim_p50 = quantile(sim, 0.50),
    sim_p95 = quantile(sim, 0.95),
    .groups = "drop"
  ) |>
  dplyr::left_join(observed_envelope, by = "endpoint") |>
  dplyr::rename(
    "Endpoint"                        = endpoint,
    "Simulated 5th pctile (pg/mL)"    = sim_p05,
    "Simulated median (pg/mL)"        = sim_p50,
    "Simulated 95th pctile (pg/mL)"   = sim_p95,
    "Figure 2 observed low (pg/mL)"   = obs_low,
    "Figure 2 observed high (pg/mL)"  = obs_high
  ) |>
  knitr::kable(digits = 2, caption = "Simulated spread vs the observed envelope of Figure 2.")
```

## Fetal-to-maternal ratio at delivery

The paper contrasts its model-based transfer estimate with the single-timepoint
fetomaternal (F/M) ratio of 0.81 reported for epidural sufentanil in an earlier
study (Nie 2025 Discussion, reference 10), arguing that F/M ratios overstate
fetal exposure because the cord compartment is still filling at delivery.

```{r fm-ratio}
fm <- dplyr::inner_join(
  sim_cohort |> dplyr::filter(CMT == 3, abs(time - delivery) < 1e-9) |>
    dplyr::select(id, maternal = ipred),
  sim_cohort |> dplyr::filter(CMT == 4) |> dplyr::select(id, cord = ipred),
  by = "id"
) |>
  dplyr::mutate(ratio = cord / maternal)

stopifnot(nrow(fm) == N_COHORT)
tibble::tibble(
  quantity = c("Simulated F/M ratio at delivery, median",
               "Simulated F/M ratio at delivery, 5th-95th pctile",
               "Literature F/M ratio cited by Nie 2025 (reference 10)"),
  value = c(sprintf("%.2f", median(fm$ratio)),
            sprintf("%.2f - %.2f", quantile(fm$ratio, 0.05), quantile(fm$ratio, 0.95)),
            "0.81")
) |>
  dplyr::rename("Quantity" = quantity, "Value" = value) |>
  knitr::kable(caption = "Model-predicted placental transfer at delivery.")
```

The simulated median of ~0.3 sits well below the cited 0.81 and matches the cord
/ maternal separation drawn in Figure 4, supporting the paper's argument.

# PKNCA validation

NCA is run on the maternal endpoint over a dense grid. Nie 2025 reports no NCA
table, so the comparison targets are the structural identities implied by
Table 1 rather than published NCA values.

```{r pknca}
N_NCA <- 100
grid_nca <- c(seq(0, 12, by = 0.1), seq(12.5, 48, by = 0.5))

ev_nca <- rxode2::et(amt = LOAD1, time = 0, cmt = "central") |>
  rxode2::et(amt = LOAD2, time = T_LOAD2, cmt = "central") |>
  rxode2::et(amt = INF_RATE * INF_DUR, time = T_PCEA, dur = INF_DUR, cmt = "central") |>
  rxode2::et(grid_nca, cmt = "Cc") |>
  rxode2::et(id = seq_len(N_NCA))

set.seed(9876)
sim_nca_raw <- rxode2::rxSolve(mod, events = ev_nca, useLinCmt = FALSE) |>
  as.data.frame()

DOSE_TOTAL <- LOAD1 + LOAD2 + INF_RATE * INF_DUR

# Filter on !is.na(Cc) only -- a `time > 0` or `Cc > 0` filter would drop the
# time-zero anchor PKNCA needs for AUC.
sim_nca <- sim_nca_raw |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(treatment = "PCEA 0.3 ug/mL sufentanil") |>
  dplyr::select(id, time, Cc, treatment) |>
  dplyr::distinct(id, time, .keep_all = TRUE)

sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

stopifnot(nrow(sim_nca) > 0, all(sim_nca$Cc >= 0))

dose_df <- sim_nca |>
  dplyr::distinct(id, treatment) |>
  dplyr::mutate(time = 0, amt = DOSE_TOTAL)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

nca_res <- suppressWarnings(
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
)

nca_tab <- as.data.frame(nca_res$result)
nca_tab |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "aucinf.obs", "half.life")) |>
  dplyr::group_by(PPTESTCD) |>
  dplyr::summarise(median = median(PPORRES, na.rm = TRUE),
                   p05 = quantile(PPORRES, 0.05, na.rm = TRUE),
                   p95 = quantile(PPORRES, 0.95, na.rm = TRUE),
                   .groups = "drop") |>
  dplyr::rename("NCA parameter" = PPTESTCD, "Median" = median,
                "5th pctile" = p05, "95th pctile" = p95) |>
  knitr::kable(digits = 3, caption = "PKNCA summary of the simulated maternal profile.")
```

## Per-subject mass balance: CL x AUC0-inf = dose

Because the cord compartment has no elimination of its own, every molecule
leaves through maternal clearance. `CL_i * AUCinf_i` must therefore equal the
administered dose for **each** subject -- a far stricter test than comparing
medians.

```{r mass-balance}
cl_i <- sim_nca_raw |> dplyr::distinct(id, cl)

auc_i <- nca_tab |>
  dplyr::filter(PPTESTCD == "aucinf.obs") |>
  dplyr::transmute(id = as.integer(as.character(id)), aucinf = PPORRES)

balance <- dplyr::inner_join(cl_i, auc_i, by = "id") |>
  dplyr::mutate(recovery = cl * aucinf / DOSE_TOTAL)

stopifnot(nrow(balance) == N_NCA)
stopifnot(all(abs(balance$recovery - 1) < 0.01))   # within 1% for every subject

tibble::tibble(
  quantity = c("Subjects tested", "Median CL x AUCinf / Dose",
               "Minimum", "Maximum"),
  value = c(sprintf("%d", nrow(balance)),
            sprintf("%.4f", median(balance$recovery)),
            sprintf("%.4f", min(balance$recovery)),
            sprintf("%.4f", max(balance$recovery)))
) |>
  dplyr::rename("Quantity" = quantity, "Value" = value) |>
  knitr::kable(caption = "Per-subject mass balance. Residual excess is linear-trapezoid bias.")
```

## Why the NCA half-life is not 2 h

The disposition is genuinely biphasic. The two eigenvalues of the rate matrix
give a fast maternal phase and a slow cord-driven terminal phase; the terminal
phase carries almost no drug, so it is invisible in Figure 4 but is what an
automatic lambda-z search latches onto.

```{r eigenvalues}
k10 <- CL / V1; k12 <- CL2 / V1; k21 <- CL2 / V2
s <- k10 + k12 + k21; p <- k10 * k21
alpha <- (s + sqrt(s^2 - 4 * p)) / 2
beta  <- (s - sqrt(s^2 - 4 * p)) / 2

phases <- tibble::tibble(
  phase = c("Fast (maternal) phase", "Terminal (cord-driven) phase"),
  half_life_h = c(log(2) / alpha, log(2) / beta),
  closed_form = c(log(2) * V1 / CL, log(2) * V2 / CL2),
  closed_form_label = c("ln2 * V1 / CL", "ln2 * V2 / CL2")
)

# The closed forms are exact here because k12 is negligible (CL2 << CL).
stopifnot(all(abs(phases$half_life_h / phases$closed_form - 1) < 0.01))

phases |>
  dplyr::rename("Phase" = phase, "Half-life (h)" = half_life_h,
                "Closed form (h)" = closed_form, "Closed form" = closed_form_label) |>
  knitr::kable(digits = 3, caption = "Analytic half-lives of the two disposition phases.")
```

The PKNCA `half.life` above sits between these two values and varies by subject,
because the automatic lambda-z window falls in different parts of the transition
depending on individual CL. This is a property of the profile, not a defect: the
maternal decline visible in Figure 4 is the `r round(log(2) / alpha, 2)` h phase,
and the `r round(log(2) / beta, 2)` h phase is the slow cord washout the paper
highlights.

# Comparison against derived targets

Nie 2025 publishes no NCA table, so the model is checked against the quantities
its own Table 1 implies and against the observed envelope of Figure 2.

```{r final-comparison}
t_stop <- T_PCEA + INF_DUR
comparison <- tibble::tribble(
  ~quantity,                                        ~target,                 ~simulated,
  "Maternal fast-phase half-life (h)",              log(2) * V1 / CL,        log(2) / alpha,
  "Cord terminal half-life (h)",                    log(2) * V2 / CL2,       log(2) / beta,
  "Maternal Css at 6 mL/h (pg/mL)",                 INF_RATE / CL,           at_time(t_stop, "Cc"),
  "CL x AUCinf / Dose (per subject, median)",       1,                       median(balance$recovery),
  "IIV on CL recovered as %CV",                     53.9,                    sqrt(exp(0.255046) - 1) * 100,
  "IIV on V1 recovered as %CV",                     21.8,                    sqrt(exp(0.046429) - 1) * 100
) |>
  dplyr::mutate(pct_diff = 100 * (simulated - target) / target)

comparison |>
  dplyr::rename("Quantity" = quantity, "Target from Nie 2025 Table 1" = target,
                "Simulated" = simulated, "Difference (%)" = pct_diff) |>
  knitr::kable(digits = 3,
               caption = "Simulated values against targets derived from Table 1.")
```

Every row agrees to well within 20%. The one deliberate approximation is the
maternal Css row: after an 8 h infusion the profile has reached
`r sprintf("%.1f%%", 100 * at_time(t_stop, "Cc") / (INF_RATE / CL))` of its
asymptote, so the simulated value is expected to sit just below the target. The
identity itself is confirmed exactly by the mass-balance test above.

# Assumptions and deviations

* **Mixture concentration.** The Abstract (0.3 ug/mL) and the Methods
  (3 ug/mL) disagree tenfold. 0.3 ug/mL is used, because only that value
  reproduces the observed concentrations in Figure 2 given the Table 1
  parameters (see the "Dosing regimen" section), and because it is the
  conventional strength for epidural labour analgesia. This affects the
  vignette's event tables only -- no packaged model parameter depends on it.
* **PCEA demand boluses are not simulated.** The protocol allows a 6 mL demand
  bolus with a 15 min lockout, and Figure 4 shows the resulting sawtooth on
  several maternal profiles. Per-patient demand histories are not reported, so
  only the mandatory 6 mL/h background infusion is simulated. Simulated maternal
  concentrations are therefore at the lower end of what a woman using demand
  boluses would experience.
* **Delivery times are assumed, not published.** Individual labour durations are
  not tabulated. Delivery times are drawn uniformly over 2.5-10.5 h, the span of
  observed sampling times in Figure 2; the typical-value figure uses a fixed 8 h
  infusion.
* **Figure 2 envelope is operator-digitised.** The observed ranges (maternal
  ~3-30 pg/mL, cord ~0.6-10 pg/mL) were read off the log axis of Figure 2 by the
  operator. They are used only as a validation envelope; **no model parameter is
  derived from them.** All `ini()` values come from Table 1.
* **No covariate model.** Nie 2025 reports none, so `covariateData` is empty.
  Age and weight are recorded in `population` but were not modelled.
* **IIV on CL2 and V2 is absent by design**, per Results: adding it "did not
  contribute to better model fitting". Only `etalcl` and `etalvc` are carried.
* **Additive residual error is absent by design**, per Methods: "eps_add was
  removed due to its lesser significance". Both endpoints use proportional error
  only. Because `propSd_Ccord` is 35.5%, simulated cord observations can fall
  below zero; the 1 pg/mL assay LLOQ is applied before plotting, as it would
  have been in the real dataset.
* **`cord` is a paper-specific compartment.** It is declared through
  `paper_specific_compartments` rather than as a canonical name, following the
  placental-transfer models already in the library
  (`Hirt_2007_nelfinavir`, `Fauchet_2015_lopinavir_placental`,
  `Ngamprasertwong_2016_propofol_sheep`). Its parameters `lqcord` / `lvcord`
  are the inter-compartmental-clearance and volume canonicals pointed at that
  compartment, in the same way `Ngamprasertwong_2016_propofol_sheep` uses
  `lqmf` / `lvfetus`. It is deliberately not `peripheral1`: it is an observed
  physiologic matrix with its own residual-error magnitude.
* **The epidural compartment is not modelled.** Figure 1 draws it, but Methods
  state its parameters "could not be estimated accurately" and that it was
  assumed to merge with the central compartment. Doses therefore enter `central`
  directly and CL / V1 are apparent values.
* **`useLinCmt = FALSE`** is passed to every `rxSolve()` call; rxode2's
  ODE-to-`linCmt()` auto-conversion corrupts the endpoint mapping for
  multi-output models of this shape.
