---
title: "Nalbuphine (Nie 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nalbuphine (Nie 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5, dpi = 100
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Nie_2023_nalbuphine"))$reference`
* Article (open access): <https://doi.org/10.3389/fphar.2023.1130287>
* Supplement: none. Frontiers in Pharmacology published no Supporting
  Information for this article, and none is required -- every final parameter
  estimate is printed in Table 4 and the final-model equations are printed
  in-line in Results section 3.2.

Nie 2023 is a single-centre prospective population PK analysis of nalbuphine
given as a single intravenous injection for induction of general anaesthesia.
Nalbuphine is a semi-synthetic opioid that agonises the kappa receptor while
partially antagonising the mu receptor, which is why it is used at induction:
it provides analgesia with a ceiling on respiratory depression.

The distinguishing feature of the model is its **single retained covariate**.
The authors screened the usual demographic and hepatic panel and found that
neither body weight nor age entered the final model; what survived
forward inclusion and backward elimination was an intra-operative quantity --
the hourly net fluid volume infused (HNF), entering intercompartmental
clearance `Q` as a power function. That negative result is the paper's main
clinical message: because weight does not scale the PK, a fixed 12 mg dose and
a weight-based 0.2 mg/kg dose give exposures that differ by less than 6%, and
the fixed regimen carries the lower variability.

```{r model-print}
mod <- readModelDb("Nie_2023_nalbuphine")
mod
```

# Population

The analysis enrolled 47 adults scheduled for elective surgery at the Fourth
Hospital of Hebei Medical University (Shijiazhuang, China) during 2021. All
received nalbuphine 15 mg intravenously over 2-3 min at induction, followed by
midazolam, sufentanil, etomidate and cisatracurium; anaesthesia was maintained
with sevoflurane and remifentanil.

458 concentrations were drawn from 48 patients. One subject (5 samples) was
dropped for a missing HNF value, leaving **27 patients / 353 samples** in the
model-building set (intensive sampling: pre-dose and 3, 5, 10, 15, 30, 45 min
and 1, 1.5, 2, 3, 4, 5, 6, 12 h) and **20 patients / 100 samples** in the
external-validation set (sparse sampling: pre-dose, during intubation, and 1,
3, 10 min after intubation). Every post-dose sample was above the 0.1 ng/mL
LLOQ and none was flagged as an outlier.

Model-building cohort baseline characteristics (Table 1), reported as
mean +/- SD (median, range):

| Characteristic | Model-building (n = 27) |
|:---|:---|
| Age (years) | 53.85 +/- 16.63 (58, 21-76) |
| Total body weight (kg) | 63.32 +/- 9.3 (62, 48-82) |
| Height (cm) | 163.96 +/- 6.87 (165, 153-175) |
| Dose (mg/kg) | 0.24 +/- 0.04 (0.24, 0.18-0.31) |
| Hourly net fluid volume infused (mL/h) | 617.96 +/- 247.61 (563.56, 234.26-1202.25) |
| Operation duration (h) | 4.31 +/- 1.9 (4, 0.83-8.33) |
| Creatinine clearance (mL/min) | 111.19 +/- 27.38 (105.14, 53.4-160.98) |
| Alanine aminotransferase (U/L) | 58.86 +/- 69.62 (23.6, 6-260.9) |
| Female | 15 (55.6%) |
| Tumour | 21 (77.8%) |
| Hepatobiliary disease | 15 (55.6%) |
| Child-Turcotte-Pugh class B | 5 (18.5%) |

Across the whole cohort, liver resection accounted for 14.8%, cholecystectomy
12.8%, pancreatic resection 36.2% and other surgery 36.2%; 29.8% of procedures
were laparoscopic and 70.2% open.

# Source trace

Every value in the model file, with the location in Nie 2023 it came from.

| Model element | Value | Source |
|:---|:---|:---|
| Structural model | 2-compartment IV | Results 3.2 ("a two-compartment model ... decreased AIC by 519.209" vs 1-compartment) |
| `lcl` | CL = 32.9 L/h | Table 4 final model; Results 3.2 printed equation `CL (L/h) = 32.9` |
| `lvc` | V1 = 32.5 L | Table 4 final model; Results 3.2 printed equation `V1 (L) = 32.5` |
| `lq` | Q = 245 L/h | Table 4 final model; Results 3.2 printed equation |
| `lvp` | V2 = 83.5 L | Table 4 final model; Results 3.2 printed equation `V2 (L) = 83.5` |
| `e_pfa_net_rate_q` | -0.58 | Table 4 final model, row "HNF on Q" |
| Q covariate form | `245 * (HNF/617.96)^-0.58` | Results 3.2 printed equation |
| HNF normalizing constant | 617.96 mL/h | Table 1, model-building cohort **mean** HNF |
| HNF definition | `(FVI + BVI - UVO) / OT` | Table 1 footnote |
| `etalcl` | omega CL = 0.277 | Table 4 final model, `omega CL` = 27.7 |
| `etalvc` | omega V1 = 0.401 | Table 4 final model, `omega V1` = 40.1 |
| `etalq` | omega Q = 0.179 | Table 4 final model, `omega Q` = 17.9 |
| `etalvp` | omega V2 = 0.408 | Table 4 final model, `omega V2` = 40.8 |
| IIV form | exponential (log-normal) | Methods 2.3.1 |
| `propSd` | 0.139 | Table 4 final model, proportional error |
| `addSd` | 2.88 ng/mL | Table 4 final model, additive error |
| Residual form | combined additive + proportional | Methods 2.3.1; Results 3.2 |
| Dosing route | IV injection into `central` | Methods 2.1 (15 mg over 2-3 min at induction) |
| Concentration units | ng/mL | Methods 2.2 (LLOQ 0.1 ng/mL, range 0.1-500 ng/mL) |

## Reading Table 4's "%CV" column

Table 4 heads the random-effects column "%CV", but the numbers printed under it
are the NONMEM omega standard deviations multiplied by 100 rather than exact
log-normal coefficients of variation. The tell is the residual-error block of
the *same* column: the final model prints the proportional error as `0.139`
while the basic model (Table 2) prints the identical quantity as `13.7`. The
software is printing the raw parameter and the header is a mislabel. Reading
the BSV rows the same way gives omega = 0.277, 0.401, 0.179 and 0.408, which is
what the model file encodes (as variances, i.e. those values squared).

This reading is corroborated by Results 3.2, which subtracts the numbers
arithmetically -- "a decrease in BSV from 35.5% to 17.9%, indicating that 17.6%
of BSV in Q was explained by HNF" (35.5 - 17.9 = 17.6). That subtraction is
only meaningful on the omega scale.

# Structural verification against a closed form

A two-compartment model dosed by IV bolus has an exact bi-exponential solution.
Building that solution from the values **printed in Table 4** -- not from the
model object's own parameters -- and comparing it to the `rxode2` ODE solve
checks the ODE wiring, the micro-constant algebra and the unit convention all
at once. A self-referential check built from `mod$theta` could not fail.

```{r closed-form}
# Values transcribed directly from Nie 2023 Table 4 / Results 3.2.
CL_pub <- 32.9   # L/h
V1_pub <- 32.5   # L
Q_pub  <- 245    # L/h at HNF = 617.96 mL/h
V2_pub <- 83.5   # L
dose_ug <- 12000 # 12 mg fixed dose from the Table 6 simulation, in ug

k10 <- CL_pub / V1_pub
k12 <- Q_pub / V1_pub
k21 <- Q_pub / V2_pub
sum_k <- k10 + k12 + k21
alpha <- (sum_k + sqrt(sum_k^2 - 4 * k10 * k21)) / 2
beta  <- (sum_k - sqrt(sum_k^2 - 4 * k10 * k21)) / 2

tgrid <- c(0, 0.05, 0.25, 1, 4, 8, 12)
closed_form <- (dose_ug / V1_pub) *
  ((alpha - k21) / (alpha - beta) * exp(-alpha * tgrid) +
     (k21 - beta) / (alpha - beta) * exp(-beta * tgrid))

ev_typ <- as.data.frame(
  rxode2::et(amt = dose_ug, cmt = "central") |> rxode2::et(tgrid)
)
ev_typ$PFA_NET_RATE <- 617.96  # the normalizing value, so Q = 245 exactly

sim_typ <- rxode2::rxSolve(
  rxode2::zeroRe(mod), ev_typ, returnType = "data.frame", addDosing = FALSE
)

data.frame(
  time = tgrid,
  closed_form = closed_form,
  ode_solve = sim_typ$Cc,
  rel_diff = sim_typ$Cc / closed_form - 1
) |>
  knitr::kable(
    caption = paste(
      "Typical-value nalbuphine concentration (ng/mL) after a 12 mg IV bolus:",
      "rxode2 ODE solve against the bi-exponential closed form built from",
      "Nie 2023 Table 4."
    ),
    digits = c(2, 4, 4, 12)
  )
```

```{r closed-form-check}
# Also confirm the solve reproduces the published half-life and that the
# derived micro-constants are what the paper's parameters imply.
stopifnot(
  # Structural gate: pure numerical-integration error, so a tight bound is
  # correct here (both sides use the same fixed parameters; nothing random).
  max(abs(sim_typ$Cc / closed_form - 1)) < 1e-6,
  # Q at the normalizing HNF must be exactly the printed 245 L/h.
  abs(sim_typ$q[1] - 245) < 1e-9,
  abs(sim_typ$cl[1] - 32.9) < 1e-9,
  abs(sim_typ$vc[1] - 32.5) < 1e-9,
  abs(sim_typ$vp[1] - 83.5) < 1e-9
)
cat(sprintf(
  "Terminal half-life = %.3f h; distribution half-life = %.4f h\n",
  log(2) / beta, log(2) / alpha
))
```

The distribution phase is extremely fast (`alpha` half-life about 3.7 min,
because Q = 245 L/h is more than seven times CL), which is why the paper's
first sample at 3 min already sits well below the back-extrapolated C0.

# The HNF covariate effect

Figure 7 of Nie 2023 shows simulated profiles at the 10th, 50th and 90th
percentiles of HNF in the model-building cohort: 350.1, 563.6 and 973.4 mL/h.
The exponent is negative, so a larger net intra-operative fluid load gives a
*smaller* intercompartmental clearance.

```{r hnf-effect}
hnf_levels <- c(`10th percentile` = 350.1, `50th percentile` = 563.6,
                `90th percentile` = 973.4)

q_expected <- 245 * (hnf_levels / 617.96)^-0.58

ev_hnf <- do.call(rbind, lapply(seq_along(hnf_levels), function(i) {
  e <- as.data.frame(
    rxode2::et(amt = 15000, cmt = "central") |>
      rxode2::et(seq(0, 12, by = 0.02))
  )
  e$id <- i
  e$PFA_NET_RATE <- hnf_levels[[i]]
  e
}))

sim_hnf <- rxode2::rxSolve(
  rxode2::zeroRe(mod), ev_hnf, returnType = "data.frame", addDosing = FALSE
)
sim_hnf$hnf_level <- factor(
  names(hnf_levels)[sim_hnf$id], levels = names(hnf_levels)
)

q_observed <- vapply(
  seq_along(hnf_levels), function(i) sim_hnf$q[sim_hnf$id == i][1], numeric(1)
)

# Exact gate: the model's q must reproduce the printed power equation to
# machine precision at every Figure 7 covariate level.
stopifnot(max(abs(q_observed / q_expected - 1)) < 1e-10)

data.frame(
  `HNF level` = names(hnf_levels),
  `HNF (mL/h)` = unname(hnf_levels),
  `Q from printed equation (L/h)` = unname(q_expected),
  `Q from model (L/h)` = q_observed,
  check.names = FALSE
) |>
  knitr::kable(
    caption = paste(
      "Intercompartmental clearance at the Figure 7 HNF percentiles.",
      "Reference values are 245 * (HNF/617.96)^-0.58 evaluated by hand."
    ),
    digits = 4
  )
```

```{r fig-hnf, fig.cap = "Replicates Figure 7 of Nie 2023: typical-value nalbuphine profiles after a 15 mg IV injection at the 10th, 50th and 90th percentiles of the hourly net fluid volume infused. As the paper reports, the covariate moves the curves very little."}
ggplot(sim_hnf, aes(time, Cc, colour = hnf_level)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)", y = "Nalbuphine concentration (ng/mL)",
    colour = "HNF level"
  ) +
  theme_bw()
```

The three curves are visually almost indistinguishable over most of the
profile. That is the paper's own conclusion: the effect is statistically
significant (dOFV -9.643, p < 0.005) and it explains a real share of the
between-subject variability in Q (35.5% down to 17.9%), but its effect on
plasma concentration is small enough that the authors explicitly recommend
against adjusting the dose for it.

# Virtual cohort

The cohort is built to match the model-building population of Table 1. Weight
is drawn from a normal distribution truncated to the reported range; HNF is
drawn from a log-normal, which reproduces the reported right skew -- the
log-normal implied by the reported mean (617.96) and SD (247.61) has a median
of about 573 mL/h against the reported 563.56.

Covariates are placed on **deterministic quantiles** rather than sampled at
random. That is the reproducible choice: it fixes the cohort's covariate
distribution exactly at the published moments, so the only randomness left in
the comparison below is the between-subject variability of the model itself.

```{r cohort}
n_sub <- 200L  # per dosing arm; well inside the 200-per-arm vignette cap
probs <- (seq_len(n_sub) - 0.5) / n_sub

wt <- pmin(pmax(qnorm(probs, mean = 63.32, sd = 9.3), 48), 82)

cv_hnf <- 247.61 / 617.96
sd_log_hnf <- sqrt(log(1 + cv_hnf^2))
hnf <- pmin(
  pmax(qlnorm(probs, log(617.96) - sd_log_hnf^2 / 2, sd_log_hnf), 234.26),
  1202.25
)

data.frame(
  Covariate = c("Total body weight (kg)", "HNF (mL/h)"),
  `Simulated mean` = c(mean(wt), mean(hnf)),
  `Published mean` = c(63.32, 617.96),
  `Simulated median` = c(median(wt), median(hnf)),
  `Published median` = c(62, 563.56),
  check.names = FALSE
) |>
  knitr::kable(
    caption = "Virtual cohort covariates against Nie 2023 Table 1.",
    digits = 2
  )
```

# Dosage regimen simulation (Table 6 and Figure 8)

Nie 2023 simulated two regimens -- a fixed 12 mg dose and a weight-based
0.2 mg/kg dose -- over the sampling grid pre-dose and 0.05, 0.08, 0.17, 0.25,
0.5, 0.75, 1, 2, 3, 4, 6, 8, 10 and 12 h, and tabulated the concentration
distribution at 0.05, 4 and 12 h (Table 6).

The two arms are simulated with **common random numbers**: `rxSetSeed()` is
called immediately before each solve so both arms draw the same etas for the
same subject. That makes the fixed-versus-weight comparison a pure dose
contrast rather than a noisy difference of two independent cohorts.

```{r regimen-sim}
t_obs <- c(0, 0.05, 0.08, 0.17, 0.25, 0.5, 0.75, 1, 2, 3, 4, 6, 8, 10, 12)

buildEvents <- function(amt_ug) {
  do.call(rbind, lapply(seq_len(n_sub), function(i) {
    e <- as.data.frame(
      rxode2::et(amt = amt_ug[[i]], cmt = "central") |> rxode2::et(t_obs)
    )
    e$id <- i
    e$PFA_NET_RATE <- hnf[[i]]
    e
  }))
}

runArm <- function(amt_ug, label) {
  ev <- buildEvents(amt_ug)
  # Reseed inside each arm so the two arms share a common random-number
  # stream; a single seed set once would give the second arm different etas.
  rxode2::rxSetSeed(20230321)
  out <- rxode2::rxSolve(mod, ev, returnType = "data.frame", addDosing = FALSE)
  out$regimen <- label
  out
}

dose_fixed <- rep(12000, n_sub)          # 12 mg in ug
dose_bw    <- 0.2 * wt * 1000            # 0.2 mg/kg in ug

sim_fixed <- runArm(dose_fixed, "Fixed 12 mg")
sim_bw    <- runArm(dose_bw,    "Bodyweight 0.2 mg/kg")
sim_reg   <- rbind(sim_fixed, sim_bw)
```

```{r regimen-linearity}
# Common-random-number gate. With identical etas and a linear (first-order)
# model, each subject's weight-based concentration must equal the fixed-dose
# concentration scaled by exactly the dose ratio, at every time point. This is
# an EXACT identity -- any dose non-linearity, covariate leakage from the dose
# amount, or eta-stream mismatch between the arms breaks it immediately.
dose_ratio <- (dose_bw / 12000)[sim_fixed$id]
conc_ratio <- sim_bw$Cc / sim_fixed$Cc
finite <- is.finite(conc_ratio) & sim_fixed$Cc > 0
stopifnot(max(abs(conc_ratio[finite] / dose_ratio[finite] - 1)) < 1e-8)
cat(sprintf(
  "Common-random-number dose linearity holds to %.2e relative error.\n",
  max(abs(conc_ratio[finite] / dose_ratio[finite] - 1))
))
```

```{r table6}
published <- tibble::tribble(
  ~time, ~regimen,               ~mean,  ~median, ~p25,   ~p75,
  0.05,  "Fixed 12 mg",          246.75, 241.84,  203.51, 284.82,
  0.05,  "Bodyweight 0.2 mg/kg", 260.22, 251.36,  206.57, 306.13,
  4,     "Fixed 12 mg",           30.14,  29.65,   23.50,  36.19,
  4,     "Bodyweight 0.2 mg/kg",  31.82,  30.78,   23.87,  38.56,
  12,    "Fixed 12 mg",            5.06,   3.99,    1.71,   7.39,
  12,    "Bodyweight 0.2 mg/kg",   5.34,   4.15,    1.76,   7.71
)

simulated <- sim_reg |>
  filter(time %in% c(0.05, 4, 12)) |>
  group_by(time, regimen) |>
  summarise(
    mean = mean(Cc), median = median(Cc),
    p25 = quantile(Cc, 0.25), p75 = quantile(Cc, 0.75),
    .groups = "drop"
  )

table6 <- published |>
  left_join(simulated, by = c("time", "regimen"),
            suffix = c("_pub", "_sim")) |>
  mutate(across(ends_with("_sim"), ~ round(.x, 2)))

table6 |>
  select(time, regimen, mean_pub, mean_sim, median_pub, median_sim,
         p25_pub, p25_sim, p75_pub, p75_sim) |>
  rename(
    "Time (h)" = time, "Regimen" = regimen,
    "Mean (pub)" = mean_pub, "Mean (sim)" = mean_sim,
    "Median (pub)" = median_pub, "Median (sim)" = median_sim,
    "25th (pub)" = p25_pub, "25th (sim)" = p25_sim,
    "75th (pub)" = p75_pub, "75th (sim)" = p75_sim
  ) |>
  knitr::kable(
    caption = paste(
      "Replicates Table 6 of Nie 2023: simulated nalbuphine concentration",
      "(ng/mL) for the two dosage regimens."
    ),
    digits = 2
  )
```

```{r table6-check}
ratios <- table6 |>
  mutate(
    r_mean = mean_sim / mean_pub, r_median = median_sim / median_pub,
    r_p25 = p25_sim / p25_pub, r_p75 = p75_sim / p75_pub
  )

early <- ratios |> filter(time %in% c(0.05, 4))
late  <- ratios |> filter(time == 12)

worst <- function(d) {
  max(abs(c(d$r_mean, d$r_median, d$r_p25, d$r_p75) - 1))
}

# Cohort gates. These compare a 200-subject virtual cohort against the
# paper's 1000-replicate simulation over its own 27 patients, so they are
# deliberately set on the CENTRE and the QUARTILES -- never on the extremes,
# whose position depends on which subjects happened to draw large etas.
#
# The 12 h row gets a looser bound on purpose: it sits 4.6 terminal
# half-lives after the dose, where the concentration spans two orders of
# magnitude across the cohort and the quartiles are governed almost entirely
# by the eta draw on CL.
stopifnot(
  worst(early) < 0.08,
  worst(late) < 0.18
)
cat(sprintf(
  "Worst |sim/pub - 1|: %.3f at 0.05-4 h, %.3f at 12 h.\n",
  worst(early), worst(late)
))
```

```{r table6-bias}
bias <- simulated |>
  select(time, regimen, mean) |>
  pivot_wider(names_from = regimen, values_from = mean) |>
  mutate(
    bias_pct = 100 * (`Bodyweight 0.2 mg/kg` / `Fixed 12 mg` - 1),
    published_bias_pct = c(5.46, 5.58, 5.55)
  )

bias |>
  rename(
    "Time (h)" = time, "Bias, simulated (%)" = bias_pct,
    "Bias, Nie 2023 Table 6 (%)" = published_bias_pct
  ) |>
  knitr::kable(
    caption = paste(
      "Bias between the two regimens. Nie 2023 reports < 6% at every time,",
      "the basis for its recommendation that a fixed dose can replace",
      "weight-based dosing."
    ),
    digits = 2
  )

# The paper's finding is that the bias stays under 6%; because body weight is
# not a model covariate, the bias is just the cohort mean dose ratio.
stopifnot(all(abs(bias$bias_pct) < 10))
```

```{r fig-regimen, fig.cap = "Replicates Figure 8 of Nie 2023: simulated nalbuphine concentration-time profiles for the fixed 12 mg regimen (left) and the 0.2 mg/kg bodyweight regimen (right). Solid line is the median, shaded band the 5th-95th percentile."}
sim_reg |>
  filter(time > 0) |>
  group_by(regimen, time) |>
  summarise(
    med = median(Cc), lo = quantile(Cc, 0.05), hi = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, med)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey70", alpha = 0.5) +
  geom_line(colour = "red", linewidth = 0.8) +
  facet_wrap(~regimen) +
  scale_y_log10() +
  labs(x = "Time after dose (h)", y = "Nalbuphine concentration (ng/mL)") +
  theme_bw()
```

The bodyweight panel is visibly wider than the fixed-dose panel, because the
weight-based regimen adds the spread of body weight on top of the model's
between-subject variability without removing any of it -- weight is not a
covariate on any disposition parameter. That is the paper's Figure 8 result and
the basis for its recommendation.

# PKNCA validation

NCA is run on the clinical 15 mg induction dose over the paper's own
model-building sampling grid (Methods 2.2). Reproducing the published
measurement grid matters: the distribution phase is so fast that a denser grid
would compute a materially different trapezoidal AUC than the study could have.

```{r nca}
nca_grid <- c(0, 3, 5, 10, 15, 30, 45) / 60
nca_grid <- c(nca_grid, 1, 1.5, 2, 3, 4, 5, 6, 12)

ev_nca <- do.call(rbind, lapply(seq_len(n_sub), function(i) {
  e <- as.data.frame(
    rxode2::et(amt = 15000, cmt = "central") |> rxode2::et(nca_grid)
  )
  e$id <- i
  e$PFA_NET_RATE <- hnf[[i]]
  e
}))

rxode2::rxSetSeed(20230321)
sim_nca_raw <- rxode2::rxSolve(
  mod, ev_nca, returnType = "data.frame", addDosing = FALSE
)
sim_nca_raw$treatment <- "15 mg IV"

conc_nca <- sim_nca_raw |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)

dose_nca <- ev_nca |>
  filter(evid == 1) |>
  select(id, time, amt) |>
  mutate(treatment = "15 mg IV")

conc_obj <- PKNCA::PKNCAconc(
  conc_nca, Cc ~ time | treatment + id, concu = "ng/mL", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  dose_nca, amt ~ time | treatment + id, doseu = "ug"
)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE,
  half.life = TRUE, cl.obs = TRUE, vz.obs = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)
```

Because doses are carried in ug and concentrations in ng/mL, PKNCA's
`cl.obs` comes out directly in L/h and `vz.obs` directly in L, with no
conversion.

## Mass-balance gate: AUC(0-inf) must equal Dose / CL

For any linear model, the area under the curve extrapolated to infinity is
exactly `Dose / CL`. Evaluated on a dense grid at the typical value, this is a
tight, deterministic check on the parameter transcription and the unit
convention together: get CL wrong, or get the ug-versus-mg convention wrong,
and it fails immediately.

```{r auc-identity}
dense <- as.data.frame(
  rxode2::et(amt = 15000, cmt = "central") |>
    rxode2::et(seq(0, 72, by = 0.002))
)
dense$PFA_NET_RATE <- 617.96

sim_dense <- rxode2::rxSolve(
  rxode2::zeroRe(mod), dense, returnType = "data.frame", addDosing = FALSE
)

auc_trapz <- sum(
  diff(sim_dense$time) *
    (head(sim_dense$Cc, -1) + tail(sim_dense$Cc, -1)) / 2
)
auc_inf <- auc_trapz + tail(sim_dense$Cc, 1) / beta
auc_expected <- 15000 / CL_pub

cat(sprintf(
  "AUC(0-inf) simulated = %.3f ng*h/mL; Dose / CL = %.3f ng*h/mL (rel. diff %.2e)\n",
  auc_inf, auc_expected, auc_inf / auc_expected - 1
))

# Deterministic identity; a tight bound is correct because both sides use the
# same fixed parameters and the only error is numerical quadrature.
stopifnot(abs(auc_inf / auc_expected - 1) < 1e-3)
```

# Comparison against published NCA

Nie 2023 does not tabulate NCA for its own cohort, but the Discussion
explicitly cross-checks the model against non-compartmental analysis of the
same patients published by the same group: "The typical values of nalbuphine PK
parameters from the final model were 32.9 L/h for CL, 32.5 L for V1, 245 L/h
for Q and 83.5 L for V2. The results were consistent with our previous results,
which were obtained using non-compartmental analysis (NCA) (Gao et al., 2022)",
which reported CL = 33.42 L/h and Vd = 137.69 L.

Those two values are the reference below. They are an *external* check: they
come from a different analysis method on the same subjects, so agreement tests
the model rather than the transcription.

```{r nca-comparison}
reference <- data.frame(
  treatment = "15 mg IV",
  cl.obs    = 33.42,   # Gao 2022 NCA, cited in Nie 2023 Introduction / Discussion
  vz.obs    = 137.69   # Gao 2022 NCA, cited in Nie 2023 Introduction / Discussion
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = reference,
  by = "treatment",
  params = c("cl.obs", "vz.obs"),
  units = c(cl.obs = "L/h", vz.obs = "L"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated NCA on the 15 mg induction dose against the non-compartmental",
    "analysis of the same cohort (Gao 2022) that Nie 2023 cites for",
    "consistency. * differs from reference by > 20%."
  ),
  align = c("l", "l", "r", "r", "r"),
  digits = 3
)
attr(cmp, "footnote")
```

```{r nca-summary}
nca_wide <- as.data.frame(nca_res) |>
  group_by(PPTESTCD) |>
  summarise(median = median(PPORRES), .groups = "drop")

nca_wide |>
  filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life",
                         "cl.obs", "vz.obs")) |>
  rename("NCA parameter" = PPTESTCD, "Simulated (median)" = median) |>
  knitr::kable(
    caption = paste(
      "Median simulated NCA parameters over the 200-subject virtual cohort",
      "after a single 15 mg IV injection."
    ),
    digits = 3
  )
```

```{r nca-check}
med <- setNames(nca_wide$median, nca_wide$PPTESTCD)

# Both NCA endpoints should land within 20% of the published NCA values.
stopifnot(
  abs(med[["cl.obs"]] / 33.42 - 1) < 0.20,
  abs(med[["vz.obs"]] / 137.69 - 1) < 0.20,
  # Terminal half-life recovered by NCA against the model's own beta. NCA
  # runs on the sparse published grid, so a few percent of disagreement is
  # expected; 20% headroom keeps this robust to which subjects draw large
  # etas without letting a real structural error through.
  abs(med[["half.life"]] / (log(2) / beta) - 1) < 0.20
)
cat(sprintf(
  "NCA cl.obs = %.2f L/h vs Gao 2022 33.42 (%.1f%%); vz.obs = %.1f L vs 137.69 (%.1f%%)\n",
  med[["cl.obs"]], 100 * (med[["cl.obs"]] / 33.42 - 1),
  med[["vz.obs"]], 100 * (med[["vz.obs"]] / 137.69 - 1)
))
```

`tmax` is 0 by construction: nalbuphine is given as an intravenous injection,
so the concentration maximum is the back-extrapolated C0 and there is no
absorption phase.

The cohort-median `aucinf.obs` lands within about 0.1% of the exact
`Dose / CL`, even though the paper's early sampling grid is coarse relative to
a 3.7-minute distribution half-life. That is PKNCA's default lin-up / log-down
integration doing its job: a linear trapezoid would over-read the convex
decline between 3 and 15 min by several percent, whereas log-linear
interpolation on the descending limb tracks it closely. The dense-grid
`Dose / CL` identity above remains the stricter, sampling-free version of the
same check.

# Assumptions and deviations

* **HNF normalizing constant.** Methods Eq. 2 states the generic power form as
  `theta1 * (cov_i / cov_median)^theta2`, but the printed final-model equation
  in Results 3.2 is `Q = 245 * (HNF/617.96)^-0.58`, and 617.96 mL/h is the
  cohort **mean** from Table 1, not the median (563.56 mL/h). The model file
  uses 617.96 per the standing rule that a printed equation outranks
  conflicting prose. The choice is not cosmetic: using the median instead
  would rescale the typical Q by `(563.56/617.96)^-0.58 = 1.055`, i.e. about
  5.5%.
* **Table 4 "%CV" column.** Read as omega standard deviations x 100 rather
  than exact log-normal CVs, for the reasons set out under "Reading Table 4's
  '%CV' column" above. Had they been exact CVs, the encoded variances would
  differ by roughly 2-4% (for example omega^2 for CL would be
  `log(1 + 0.277^2) = 0.0739` instead of `0.277^2 = 0.0767`) -- small, but the
  evidence for the standard-deviation reading is direct.
* **Unit convention.** Doses are expressed in ug so that `central / vc` yields
  ng/mL, the unit the paper reports. The 15 mg clinical dose is therefore
  15000 ug. This avoids rewriting the published additive residual error
  (2.88 ng/mL) as a converted value.
* **Table 4's "Q (L)" line.** The printed final-model equation labels Q with
  the unit "L". This is a typo: Table 4's legend, the Discussion and
  dimensional consistency all give Q in L/h. Encoded as L/h.
* **Bolus versus 2-3 min injection.** The paper administered nalbuphine over
  2-3 min but its own Table 6 simulation is reproduced almost exactly by a
  bolus (typical value 248.1 ng/mL at 0.05 h against a published mean of
  246.75), so the simulations here use a bolus. Users who need the injection
  duration can supply `dur` or `rate` on the dose record.
* **Screened but unretained covariates.** ALT, GGT, HR, WT, UA, D-dimer,
  CTP class, cancer status, operation duration and smoking status were all
  tested (Table 3) and none is in the final model. They are recorded in the
  model file's `covariatesDataExcluded` rather than `covariateData`, because
  the paper prints no point estimate for any of them -- only the OFV drops --
  so none can be reconstructed. ALT is the near-miss: it entered the full
  model on CL but its backward-elimination dOFV of 7.40 fell short of the 7.88
  retention criterion.
* **Virtual-cohort distributions.** Table 1 reports mean, SD, median and range
  but not the distributional family. Weight is assumed normal truncated to the
  reported range and HNF log-normal (which reproduces the reported right skew
  to within 2% on the median). No correlation between weight and HNF is
  imposed; the paper reports the correlation matrix only graphically
  (Figure 3) and does not tabulate a weight-HNF coefficient.
* **Cohort size.** The paper simulated 1000 replicates over its 27 patients;
  this vignette uses a 200-subject cohort per arm, the library's per-arm cap.
  The Table 6 comparison is therefore made on the centre and the quartiles,
  with a looser tolerance at 12 h where the cohort spread is widest.
* **Race and ethnicity.** Not reported. The cohort is single-centre Chinese;
  no race covariate exists in the model.
* **External-validation dataset.** Not reproduced here. Table 5's MDPE / MAPE /
  F20 / F30 statistics are computed against the 20-patient observed
  concentrations, which are not published, so they cannot be recomputed from
  the model alone.
