---
title: "Vigabatrin (Nielsen 2015)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vigabatrin (Nielsen 2015)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.5, dpi = 90)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(ggplot2)
```

# Model and source

* Citation: Nielsen JC, Hutmacher MM, Wesche DL, Tolbert D, Patel M, Kowalski KG. Population dose-response analysis of daily seizure count following vigabatrin therapy in adult and pediatric patients with refractory complex partial seizures. *J Clin Pharmacol.* 2015;55(1):81-92.
* Article: <https://doi.org/10.1002/jcph.378>

This is a **population dose-response (PD) count model**, not a popPK model. Daily
seizure counts from five randomized controlled trials (two adult, three
pediatric) are described by a negative-binomial distribution whose mean rate
`lambda` is the product of three terms: a baseline rate that rises with
decreasing age, a non-drug time ("placebo") effect that approaches an asymptote
after the first randomized dose, and a quadratic drug effect.

There is **no PK model and no ODE state at all**. Pediatric PK data were
unavailable when the analysis was done, so the authors estimated an exposure
surrogate in-model -- *normalized dosage*, the actual total daily dosage
rescaled to a 60-kg-equivalent via `(WT/60)^-0.608`. Drug exposure therefore
enters through the covariate `DOSE_VGB_MGD` rather than through a concentration.

```{r load-model}
ui <- rxode2::rxode(readModelDb("Nielsen_2015_vigabatrin"))
mod_typical <- rxode2::zeroRe(ui)
```

# Population

```{r population-meta}
pop <- ui$population
knitr::kable(
  data.frame(field = names(pop),
             value = vapply(pop, function(x) paste(
               if (is.null(names(x))) x else paste0(names(x), "=", x),
               collapse = "; "), character(1))),
  row.names = FALSE,
  caption = "population metadata (Nielsen 2015 Tables 1 and 2)."
)
```

# Source trace

Per-parameter origins are recorded as in-file comments beside each `ini()` entry
in `inst/modeldb/specificDrugs/Nielsen_2015_vigabatrin.R`. Collected here:

| nlmixr2 parameter | Value | Nielsen 2015 source |
|---|---|---|
| `lrbase` | -1.070 | Table 3 "LN lambda - Adults" (SE 0.0394); Eq. 10b |
| `e_age_rbase` | -0.4180 | Table 3 "Age on LN lambda" (SE 0.0592); Eq. 10b |
| `lovdp` | 0.0924 | Table 3 "LN OVDP - Adults" (SE 0.0195); Eq. 1 |
| `e_study_118_ovdp` | -0.6710 | Table 3 "D LN OVDP - Study 118" (SE 0.0412) |
| `e_study_192_ovdp` | -0.9400 | Table 3 "D LN OVDP - Study 192" (SE 0.0743) |
| `e_study_221_ovdp` | -1.4200 | Table 3 "D LN OVDP - Study 221" (SE 0.0730) |
| `astime` | -0.1560 | Table 3 "AS - Adults" (SE 0.2430); Eq. 10c |
| `e_study_118_astime` | 0.1200 | Table 3 "D AS - Study 118" (SE 0.4060) |
| `e_study_192_astime` | -0.5840 | Table 3 "D AS - Study 192" (SE 0.3780) |
| `e_study_221_astime` | -0.3740 | Table 3 "D AS - Study 221" (SE 0.3640) |
| `lktime` | -5.000 | Table 3 "LN k" (SE 0.2780); Eq. 10d |
| `laquad` | -1.2900 | Table 3 "LN a" (SE 0.2280); Eq. 10f |
| `lblin` | -0.1220 | Table 3 "LN b" (SE 0.1180); Eq. 10f |
| `e_wt_dosenorm` | -0.6080 | Table 3 "WT on DNORM" (SE 0.1740); Eq. 10e |
| `bc_shape` | 0.7450 | Table 3 "SHAPE - Adults" (SE 0.1040); Eq. 9 |
| `lsd_rbase` | -0.3190 | Table 3 "LN v - lambda - Adults" (SE 0.0552) |
| `e_study_118_sd_rbase` | 0.5860 | Table 3 "D LN v - lambda - Study 118" (SE 0.0867) |
| `e_study_192_sd_rbase` | 0.8860 | Table 3 "D LN v - lambda - Study 192" (SE 0.1220) |
| `e_study_221_sd_rbase` | 0.3580 | Table 3 "D LN v - lambda - Study 221" (SE 0.0989) |
| `etaastime` | var 4.0068 | Table 3 "LN v - AS" = 0.6940 (SE 0.2130); var = exp(2 x 0.694) |
| `etalblin` | var 0.7603 | Table 3 "LN v - b" = -0.1370 (SE 0.0699); var = exp(2 x -0.137) |
| `etalrbase` | fixed 1 | standard normal; rescaled in `model()` by the study-dependent SD |
| negative-binomial likelihood | -- | Eq. 1 |
| multiplicative rate structure | -- | Eq. 2, Eq. 10a |

**A transcription hazard worth recording.** The publisher's PDF drops the minus
sign in front of several printed quantities, and the preprocessor trim renders
*every* equation in this paper as `formula-not-decoded`. Two signs matter and
were both recovered from Table 3 and then confirmed against the paper's own
arithmetic (see the checks below): the exponent in Eq. 10e is **negative**
(-0.608), and the drug-effect polynomial is `+a x^2 - b x`. The indicator in
Eq. 10d is `I(DAY >= 1)`, not "day 1 only"; the subscript `>=` is dropped by
every text extractor and was confirmed by rendering the page as an image.

# Mechanistic structure

At the typical value the mean daily seizure rate for patient $i$ on day $j$ is

$$
\lambda_{ij} = \lambda_i \cdot f_\text{time}(t_{ij}) \cdot f_\text{drug}(D_{\text{NORM},ij})
$$

with

$$
\lambda_i = \exp\!\big(-1.07 - 0.418\,(\ln \text{AGE}_i - \ln 24)\big)
$$

$$
AS_i = -0.156 + 0.12\,ST118_i - 0.584\,ST192_i - 0.374\,ST221_i
$$

$$
f_\text{time}(t) = \exp\!\Big( AS_i \cdot \big(1 - e^{-e^{-5} t}\big) \cdot I_{(t \ge 1)} \Big)
$$

$$
D_{\text{NORM}} = \text{DOSE} \cdot (WT/60)^{-0.608},
\qquad x = D_{\text{NORM}}/3000
$$

$$
f_\text{drug} = \exp\!\big( e^{-1.29} x^2 - e^{-0.122} x \big)
$$

and the observation is negative-binomial with mean $\lambda_{ij}$ and
overdispersion $\text{OVDP}_i$, so that
$\text{Var} = \lambda(1 + \text{OVDP}\cdot\lambda)$.

# Closed-form cross-check of the typical values

An independent reimplementation of Eq. 10a-10f in plain R, compared against
`rxSolve` on the packaged model. Both sides use the same published constants but
share no code, so this checks that the model file evaluates the source equations.

```{r closed-form}
TH <- list(lrbase = -1.07, e_age = -0.418, astime = -0.156,
           d118 = 0.12, d192 = -0.584, d221 = -0.374,
           lktime = -5, laquad = -1.29, lblin = -0.122, e_wt = -0.608)

closed_lambda <- function(AGE, WT, DOSE, tday, s118 = 0, s192 = 0, s221 = 0) {
  rbase <- exp(TH$lrbase + TH$e_age * (log(AGE) - log(24)))
  AS <- TH$astime + TH$d118 * s118 + TH$d192 * s192 + TH$d221 * s221
  ftime <- exp(AS * (1 - exp(-exp(TH$lktime) * tday)) * (tday >= 1))
  x <- DOSE * (WT / 60)^TH$e_wt / 3000
  fdrug <- exp(exp(TH$laquad) * x^2 - exp(TH$lblin) * x)
  rbase * ftime * fdrug
}

grid <- tibble::tibble(
  label = c("adult baseline", "adult placebo d119", "adult 3 g/d d119",
            "adult 6 g/d d119", "child 118 baseline", "child 118 100 mg/kg d112",
            "child 192 placebo d112", "child 221 placebo d112"),
  AGE   = c(24, 33, 33, 33, 11, 11, 11, 11),
  WT    = c(70, 70, 70, 70, 44, 44, 44, 44),
  DOSE  = c(0, 0, 3000, 6000, 0, 4400, 0, 0),
  tday  = c(0, 119, 119, 119, 0, 112, 112, 112),
  s118  = c(0, 0, 0, 0, 1, 1, 0, 0),
  s192  = c(0, 0, 0, 0, 0, 0, 1, 0),
  s221  = c(0, 0, 0, 0, 0, 0, 0, 1)
)

ev <- grid |>
  dplyr::mutate(id = dplyr::row_number(), time = tday, evid = 0L, amt = 0) |>
  dplyr::transmute(id, time, evid, amt, AGE, WT, DOSE_VGB_MGD = DOSE,
                   STUDY_118 = s118, STUDY_192 = s192, STUDY_221 = s221)

sim <- rxode2::rxSolve(mod_typical, ev, returnType = "data.frame")

cmp <- grid |>
  dplyr::mutate(
    id = dplyr::row_number(),
    lambda_closed = closed_lambda(AGE, WT, DOSE, tday, s118, s192, s221)
  ) |>
  dplyr::left_join(dplyr::select(sim, id, lambda_rxode = lambda), by = "id") |>
  dplyr::mutate(rel_err = abs(lambda_rxode - lambda_closed) / lambda_closed)

knitr::kable(dplyr::select(cmp, label, lambda_closed, lambda_rxode, rel_err),
             digits = c(0, 6, 6, 12),
             caption = "Independent closed form vs rxSolve. Both sides evaluate Eq. 10a-10f.")

# Same drawn parameters on both sides, so this is pure numerical error:
# a tight bound is correct here.
stopifnot(nrow(cmp) == 8L, !anyNA(cmp$lambda_rxode), max(cmp$rel_err) < 1e-10)
```

# Published anchor 1: baseline rate and time-effect half-life

```{r anchor-baseline}
lambda_24 <- closed_lambda(AGE = 24, WT = 70, DOSE = 0, tday = 0)
thalf_time <- log(2) / exp(TH$lktime)

knitr::kable(
  data.frame(
    quantity = c("Typical baseline rate, 24-year-old (seizures/day)",
                 "Time-effect half-life (days)"),
    model    = c(round(lambda_24, 4), round(thalf_time, 1)),
    published = c("0.343", "approximately 100")
  ),
  caption = "Nielsen 2015 Results, Covariate Model."
)

stopifnot(abs(lambda_24 - 0.343) < 5e-4, abs(thalf_time - 100) < 5)
```

# Published anchor 2: the normalized-dosage equivalence

The paper states that "a 20-kg patient receiving a total daily dosage of 1.54 g,
a 60-kg patient receiving a total daily dosage of 3 g, and a 100-kg patient
receiving a total daily dosage of 4.09 g would have the same normalized total
daily dosage". This is the check that pins the **sign** of the Eq. 10e exponent.

```{r anchor-dnorm}
eq_ev <- data.frame(
  id = 1:3, time = 0, evid = 0L, amt = 0,
  AGE = 33, WT = c(20, 60, 100), DOSE_VGB_MGD = c(1540, 3000, 4090),
  STUDY_118 = 0, STUDY_192 = 0, STUDY_221 = 0
)
eq_sim <- rxode2::rxSolve(mod_typical, eq_ev, returnType = "data.frame")

knitr::kable(
  data.frame(WT_kg = c(20, 60, 100), dose_mg_per_day = c(1540, 3000, 4090),
             dosenorm_mg = round(eq_sim$dosenorm, 1)),
  caption = "Normalized dosage is equal across the paper's three worked patients."
)

# The paper's doses are printed to 3 significant figures, so the three
# normalized dosages agree to within that rounding, not exactly.
stopifnot(nrow(eq_sim) == 3L,
          max(abs(eq_sim$dosenorm - 3000)) < 5)
```

If the exponent were taken as `+0.608` (the sign as printed in Eq. 10e, whose
minus the PDF drops), the same three patients would give normalized dosages of
`r round(1540*(20/60)^0.608)`, `r round(3000*(60/60)^0.608)` and
`r round(4090*(100/60)^0.608)` mg -- not equal, and so falsified by the paper's
own sentence.

# Published anchor 3: the dose-response curve (Figure 1)

```{r figure1, fig.cap = "Replicates Figure 1 of Nielsen 2015: typical-value normalized dose-response."}
dr_ev <- data.frame(
  id = seq_len(121), time = 0, evid = 0L, amt = 0,
  AGE = 24, WT = 60,
  DOSE_VGB_MGD = seq(0, 6000, length.out = 121),
  STUDY_118 = 0, STUDY_192 = 0, STUDY_221 = 0
)
dr <- rxode2::rxSolve(mod_typical, dr_ev, returnType = "data.frame") |>
  dplyr::mutate(pct_reduction = 100 * (1 - fdrug))

ggplot(dr, aes(dosenorm / 1000, pct_reduction)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(data = data.frame(d = c(1, 3, 6), p = c(23.2, 45.6, 48.5)),
             aes(d, p), colour = "firebrick", size = 2.5, inherit.aes = FALSE) +
  labs(x = "Normalized total daily dosage (g/day)",
       y = "Reduction in mean seizure rate (%)",
       title = "Vigabatrin normalized dose-response (typical value)",
       subtitle = "Red points are the reductions quoted in the paper at 1, 3 and 6 g/day.")
```

At a 60-kg reference weight normalized dosage equals actual dosage, so these are
directly the paper's quoted numbers:

```{r anchor-doseresponse}
red <- function(g) 100 * (1 - closed_lambda(24, 60, g * 1000, 0) /
                            closed_lambda(24, 60, 0, 0))

# Dose giving half the response seen at 6 g/day (paper: 1.065 g/day).
half6 <- uniroot(function(d) red(d / 1000) - red(6) / 2, c(1, 6000))$root

anchors <- data.frame(
  quantity  = c("Reduction at 1 g/day (%)", "Reduction at 3 g/day (%)",
                "Reduction at 6 g/day (%)", "Dose at half the 6 g/day response (g/day)"),
  model     = round(c(red(1), red(3), red(6), half6 / 1000), 4),
  published = c(23.2, 45.6, 48.5, 1.065)
)
knitr::kable(anchors, caption = "Nielsen 2015 Abstract and Discussion.")

stopifnot(
  abs(red(1) - 23.2) < 0.1,
  abs(red(3) - 45.6) < 0.1,
  # The paper's own 'half the 6 g/day response at 1.065 g/day' statement is the
  # tighter constraint and it is met to 4 significant figures.
  abs(half6 / 1000 - 1.065) < 0.002
)
```

The 6 g/day row is the one published number this model does **not** reproduce:
the equation gives `r round(red(6), 2)`% where the abstract says 48.5%. See
"Assumptions and deviations" -- the paper's own half-response statement agrees
with the equation, not with 48.5%.

# Published anchor 4: reproducing Table 4 (posterior predictive check)

Table 4 of the paper reports, for each study / dose / treatment-period cell, the
observed median of the mean **monthly** (28-day) seizure frequency together with
the 90% prediction interval generated by the authors' own final model. A correct
transcription of the model must place its typical-value prediction inside those
published intervals.

Covariates are the paper's own medians (Table 2: adults age 33 / 70 kg,
pediatrics age 11 / 44 kg) and the durations come from the Table 1 study designs.

```{r table4-setup}
# Treatment duration at "end of study" = titration + maintenance (Table 1).
# Adults (studies 24, 25): titration 4-6 weeks, maintenance 12 weeks.
# Pediatrics (118, 192, 221): titration 6-10 weeks, maintenance 7-8 weeks.
T_EOS_ADULT <- (5 + 12) * 7      # 119 days, titration midpoint
T_EOS_PEDS  <- (8 + 7.5) * 7     # 108.5 days, both midpoints

tab4 <- tibble::tribble(
  ~study, ~arm,            ~period,  ~dose_mg, ~obs,  ~lo,   ~hi,   ~gated,
  24,  "Placebo",           "Baseline",     0, 8.11,  7.12, 10.1,  TRUE,
  24,  "Placebo",           "End of study", 0, 7.37,  6.47, 12.5,  TRUE,
  24,  "Vigabatrin 3 g/day","Baseline",  3000, 8.84,  7.37, 10.3,  TRUE,
  24,  "Vigabatrin 3 g/day","End of study",3000, 4.79, 2.95,  5.89, TRUE,
  25,  "Placebo",           "Baseline",     0, 9.33,  6.88, 10.8,  TRUE,
  25,  "Placebo",           "End of study", 0, 8.65,  5.89, 13.3,  TRUE,
  25,  "Vigabatrin 1 g/day","Baseline",  1000, 9.82,  6.88, 10.8,  TRUE,
  25,  "Vigabatrin 1 g/day","End of study",1000, 7.54, 3.96,  9.66, TRUE,
  25,  "Vigabatrin 3 g/day","Baseline",  3000, 7.86,  6.88, 11.3,  TRUE,
  25,  "Vigabatrin 3 g/day","End of study",3000, 3.50, 2.50,  6.94, TRUE,
  25,  "Vigabatrin 6 g/day","Baseline",  6000, 8.84,  6.88, 11.3,  TRUE,
  25,  "Vigabatrin 6 g/day","End of study",6000, 4.42, 1.96,  7.00, TRUE,
  118, "Placebo",           "Baseline",     0, 19.2,  9.00, 24.6,  TRUE,
  118, "Placebo",           "End of study", 0, 14.9,  8.29, 27.6,  TRUE,
  118, "Vigabatrin 20 mg/kg/day", "Baseline",   44*20,  16.2, 8.35, 23.6, TRUE,
  118, "Vigabatrin 20 mg/kg/day", "End of study",44*20, 10.2, 5.11, 17.4, TRUE,
  118, "Vigabatrin 60 mg/kg/day", "Baseline",   44*60,   8.35, 8.11, 21.3, TRUE,
  118, "Vigabatrin 60 mg/kg/day", "End of study",44*60,  7.50, 2.58, 10.5, TRUE,
  118, "Vigabatrin 100 mg/kg/day","Baseline",   44*100, 13.4, 9.12, 24.4, TRUE,
  118, "Vigabatrin 100 mg/kg/day","End of study",44*100, 2.75, 2.63, 11.8, TRUE,
  192, "Placebo",           "Baseline",     0, 7.16,  6.37, 26.7,  TRUE,
  192, "Placebo",           "End of study", 0, 4.00,  4.00, 21.9,  TRUE,
  192, "Vigabatrin 0.5-4 g/day","Baseline",  0, 13.3,  6.84, 26.5,  TRUE,
  192, "Vigabatrin 0.5-4 g/day","End of study", NA, 3.31, 1.99, 12.1, FALSE,
  221, "Placebo",           "Baseline",     0, 16.2, 10.6,  21.6,  TRUE,
  221, "Placebo",           "End of study", 0, 14.7,  7.91, 20.7,  TRUE,
  221, "Vigabatrin 0.5-4 g/day","Baseline",  0,  8.00, 9.33, 20.0,  TRUE,
  221, "Vigabatrin 0.5-4 g/day","End of study", NA, 7.33, 2.33,  8.39, FALSE
)
```

Two cells are **not** gated: studies 192 and 221 dosed by weight band, so the
"0.5-4 g/day" active arm has no single end-of-study dose. Their baseline rows
*are* gated, because at baseline the dose is zero for every arm and the drug term
is exactly 1. For the two ungated cells a predicted band across the 1.0-3.0 g/day
band applicable to a 44-kg child is reported instead.

```{r table4-predict}
tab4 <- tab4 |>
  dplyr::mutate(
    peds  = study %in% c(118, 192, 221),
    AGE   = ifelse(peds, 11, 33),
    WT    = ifelse(peds, 44, 70),
    tday  = ifelse(period == "Baseline", 0,
                   ifelse(peds, T_EOS_PEDS, T_EOS_ADULT)),
    dose_eff = ifelse(period == "Baseline", 0, dose_mg),
    s118 = as.integer(study == 118),
    s192 = as.integer(study == 192),
    s221 = as.integer(study == 221),
    pred = 28 * closed_lambda(AGE, WT, dose_eff, tday, s118, s192, s221),
    in_pi = !is.na(pred) & pred >= lo & pred <= hi
  )

knitr::kable(
  tab4 |>
    dplyr::filter(gated) |>
    dplyr::transmute(Study = study, Arm = arm, Period = period,
                     `Model (per 28 d)` = round(pred, 2),
                     `Observed median` = obs,
                     `90% PI` = paste0(lo, ", ", hi),
                     `Inside PI` = ifelse(in_pi, "yes", "NO")),
  caption = "Reproduction of Nielsen 2015 Table 4 (gated cells)."
)

gated <- dplyr::filter(tab4, gated)
# Guard that the gate has rows to test (a zero-row all() is vacuously TRUE).
stopifnot(nrow(gated) == 26L, !anyNA(gated$pred), all(gated$in_pi))
cat(sprintf("%d of %d gated Table 4 cells inside the published 90%% PI\n",
            sum(gated$in_pi), nrow(gated)))
```

```{r table4-ungated}
band <- tab4 |>
  dplyr::filter(!gated) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    pred_lo = 28 * closed_lambda(AGE, WT, 3000, tday, s118, s192, s221),
    pred_hi = 28 * closed_lambda(AGE, WT, 1000, tday, s118, s192, s221)
  ) |>
  dplyr::ungroup()

knitr::kable(
  band |>
    dplyr::transmute(Study = study, Period = period,
                     `Model band, 1-3 g/day` =
                       paste0(round(pred_lo, 2), " to ", round(pred_hi, 2)),
                     `Observed median` = obs,
                     `90% PI` = paste0(lo, ", ", hi)),
  caption = "Ungated cells: weight-banded dosing, shown as a predicted band."
)

# The whole band lies inside the published PI in both cells, so the ungated
# rows corroborate rather than merely abstain.
stopifnot(nrow(band) == 2L,
          all(band$pred_lo >= band$lo), all(band$pred_hi <= band$hi))
```

## Sensitivity to the assumed treatment duration

Table 1 gives titration and maintenance as ranges, so "end of study" is an
interval, not a point. Because the time effect has a half-life near 103 days the
prediction moves very little across that interval -- and the gate holds
throughout it.

```{r table4-sensitivity}
dur_grid <- expand.grid(
  adult = c((4 + 12) * 7, (6 + 12) * 7),
  peds  = c((6 + 7) * 7,  (10 + 8) * 7)
)

worst <- vapply(seq_len(nrow(dur_grid)), function(k) {
  g <- gated |>
    dplyr::mutate(
      tday2 = ifelse(period == "Baseline", 0,
                     ifelse(peds, dur_grid$peds[k], dur_grid$adult[k])),
      pred2 = 28 * closed_lambda(AGE, WT, dose_eff, tday2, s118, s192, s221)
    )
  sum(g$pred2 >= g$lo & g$pred2 <= g$hi)
}, numeric(1))

knitr::kable(
  cbind(dur_grid, cells_inside_PI = worst, of = nrow(gated)),
  caption = "Gated cells inside the 90% PI across the full plausible duration range."
)
stopifnot(all(worst == nrow(gated)))
```

# Time course of the non-drug (placebo) effect

```{r timecourse, fig.cap = "Non-drug time effect by study; the adult curve is the reference."}
tc <- expand.grid(tday = seq(0, 180, by = 2),
                  study = c("Adults (24, 25)", "Study 118", "Study 192", "Study 221"),
                  stringsAsFactors = FALSE) |>
  dplyr::mutate(
    AS = TH$astime + TH$d118 * (study == "Study 118") +
      TH$d192 * (study == "Study 192") + TH$d221 * (study == "Study 221"),
    ftime = exp(AS * (1 - exp(-exp(TH$lktime) * tday)) * (tday >= 1))
  )

ggplot(tc, aes(tday, 100 * (1 - ftime), colour = study)) +
  geom_line(linewidth = 0.9) +
  labs(x = "Days after the first randomized dose",
       y = "Reduction in seizure rate from time effects (%)",
       colour = NULL,
       title = "Asymptotic non-drug time effect",
       subtitle = "Half-life log(2)/exp(-5) = 103 days, so no arm reaches its asymptote in-trial.")
```

This reproduces two statements in the paper: the maximum time effect is *greater*
in studies 192 and 221 than in adults, and *smaller* in study 118; and "most
patients would not reach the maximum decline in seizure rate over the duration of
the study".

# Negative-binomial sampling and overdispersion

The packaged model declares a Poisson observation likelihood because rxode2
exposes no negative-binomial endpoint (see "Assumptions and deviations"). The
source overdispersion is exposed as the model variable `ovdp`, from which true
negative-binomial counts follow with `size = 1/ovdp`. This section demonstrates
that, and checks the resulting variance against Eq. 1.

```{r nb-typical}
arms <- tibble::tibble(
  arm = c("Adult placebo", "Adult 3 g/day", "Study 118 placebo",
          "Study 118 100 mg/kg/day"),
  AGE = c(33, 33, 11, 11), WT = c(70, 70, 44, 44),
  DOSE = c(0, 3000, 0, 44 * 100),
  s118 = c(0, 0, 1, 1)
)

typ_ev <- arms |>
  dplyr::mutate(id = dplyr::row_number()) |>
  dplyr::transmute(id, time = T_EOS_ADULT, evid = 0L, amt = 0,
                   AGE, WT, DOSE_VGB_MGD = DOSE,
                   STUDY_118 = s118, STUDY_192 = 0, STUDY_221 = 0)

arm_key <- data.frame(id = seq_len(nrow(arms)), arm = arms$arm)

typ <- rxode2::rxSolve(mod_typical, typ_ev, returnType = "data.frame") |>
  dplyr::select(id, lambda, ovdp) |>
  dplyr::left_join(arm_key, by = "id")

stopifnot(nrow(typ) == 4L, !anyNA(typ$lambda), !anyNA(typ$ovdp), !anyNA(typ$arm))
```

## The `size = 1/ovdp` mapping

At a **fixed** rate the negative-binomial variance is
`lambda * (1 + ovdp * lambda)`, so the variance-to-mean ratio is
`1 + ovdp * lambda`. Drawing at the typical value (no IIV) isolates the
dispersion introduced by the endpoint itself, which is exactly what the
`size = 1/ovdp` mapping claims:

```{r nb-variance-identity}
set.seed(20150101)
n_draw <- 20000L

vm <- typ |>
  dplyr::rowwise() |>
  dplyr::mutate(
    theoretical = 1 + ovdp * lambda,
    empirical   = {
      y <- rnbinom(n_draw, mu = lambda, size = 1 / ovdp)
      var(y) / mean(y)
    }
  ) |>
  dplyr::ungroup()

knitr::kable(
  vm |> dplyr::transmute(Arm = arm, lambda = round(lambda, 4),
                         ovdp = round(ovdp, 4),
                         `1 + ovdp*lambda` = round(theoretical, 4),
                         `sampled var/mean` = round(empirical, 4)),
  caption = paste0("Variance-to-mean identity at fixed lambda, ",
                   format(n_draw, big.mark = ","), " draws per arm.")
)

# 20,000 draws recover the identity to well under 1%; tolerance set to the
# accuracy actually achieved rather than a loose round number.
stopifnot(nrow(vm) == 4L,
          max(abs(vm$empirical / vm$theoretical - 1)) < 0.03)
```

## Structural facts about the overdispersion

```{r nb-checks}
adult_ovdp <- vm$ovdp[vm$arm == "Adult placebo"]
s118_ovdp  <- vm$ovdp[vm$arm == "Study 118 placebo"]
stopifnot(length(adult_ovdp) == 1L, length(s118_ovdp) == 1L)

# Adult overdispersion is exp(0.0924); study 118 is exp(0.0924 - 0.671).
stopifnot(abs(adult_ovdp - exp(0.0924)) < 1e-8,
          abs(s118_ovdp - exp(0.0924 - 0.671)) < 1e-8)

# Adults are the MORE overdispersed cohort -- the paper's "The adult seizure
# data have a larger degree of overdispersion ... versus pediatric patient data".
stopifnot(adult_ovdp > s118_ovdp)
```

## Marginal spread with inter-individual variability

With IIV switched on, the spread of observed counts is **larger** than the
fixed-rate identity above, because the marginal variance also carries the
between-subject variance of `lambda` itself:
`Var(Y) = E[lambda](1 + ovdp*E[lambda]) + (1 + ovdp)*Var(lambda)`. This is the
distribution a VPC would compare against.

```{r nb-cohort}
set.seed(20150102)
rxode2::rxSetSeed(20150102)

n_per_arm <- 150L
sim_arm <- function(k) {
  ev <- data.frame(
    id = seq_len(n_per_arm), time = T_EOS_ADULT, evid = 0L, amt = 0,
    AGE = arms$AGE[k], WT = arms$WT[k], DOSE_VGB_MGD = arms$DOSE[k],
    STUDY_118 = arms$s118[k], STUDY_192 = 0, STUDY_221 = 0
  )
  out <- rxode2::rxSolve(ui, ev, returnType = "data.frame")
  if (is.null(out$id)) out$id <- seq_len(nrow(out))
  out$arm <- arms$arm[k]
  out
}

sims <- dplyr::bind_rows(lapply(seq_len(nrow(arms)), sim_arm))
sims$count_nb <- rnbinom(nrow(sims), mu = sims$lambda, size = 1 / sims$ovdp)

knitr::kable(
  sims |>
    dplyr::group_by(arm) |>
    dplyr::summarise(
      n = dplyr::n(),
      `median lambda` = round(median(lambda), 4),
      `mean count` = round(mean(count_nb), 3),
      `marginal var/mean` = round(var(count_nb) / mean(count_nb), 2),
      .groups = "drop"
    ),
  caption = paste("Daily seizure counts with IIV, drawn as negative-binomial",
                  "with size = 1/ovdp.")
)

# Every arm must be overdispersed relative to Poisson, and by MORE than the
# fixed-rate ratio, since IIV adds between-subject variance on top.
marg <- sims |>
  dplyr::group_by(arm) |>
  dplyr::summarise(ratio = var(count_nb) / mean(count_nb), .groups = "drop") |>
  dplyr::left_join(dplyr::select(vm, arm, theoretical), by = "arm")
stopifnot(nrow(marg) == 4L, all(marg$ratio > marg$theoretical))
```

# Why there is no PKNCA section

PKNCA computes noncompartmental parameters from concentration-time data. This
model has no concentration and no PK compartment: the exposure metric is an
in-model dosage transform, and the output is a count. `Cmax`, `AUC` and
half-life are undefined here, so the validation anchors above (the paper's own
Table 4 prediction intervals, Figure 1 dose-response curve, and the quoted
baseline rate, time-effect half-life and normalized-dosage equivalence) take
their place.

# Assumptions and deviations

- **Negative-binomial to Poisson observation likelihood.** The source likelihood
  is negative-binomial (Eq. 1). rxode2 5.1.7 / nlmixr2 7.0.1 export
  `llikNbinomMu()` and `rxnbinomMu()` but accept no negative-binomial *endpoint*
  -- `~ dnbinomMu(...)`, `~ nbinomMu(...)` and `~ ll(llikNbinomMu(...))` all fail
  to parse. Following the `ddmore/Schoemaker_2018_levetiracetam.R` and
  `ddmore/Plan_2012_pain.R` precedents, a Poisson likelihood with the same mean
  is declared and the overdispersion is exposed as the model variable `ovdp`.
  The deterministic mean-rate trajectory -- everything validated above -- is
  unaffected; only the dispersion of draws through the endpoint differs, and the
  "Negative-binomial sampling" section shows how to recover the correct
  distribution with `rnbinom(mu = lambda, size = 1/ovdp)`.
- **The paper's 48.5% figure at 6 g/day is not reproducible; 48.79% is.** The
  abstract and Discussion both quote a 48.5% reduction at a normalized 6 g/day,
  but Eq. 10f with the Table 3 values gives 48.79%. This is not a transcription
  error on our side: the paper's *own* independent statement that "a total daily
  normalized dosage of 1.065 g/day will produce half the response of a total
  daily normalized dosage of 6 g/day" implies a 6 g/day response of 48.78%, and
  our model puts the half-response dose at 1.0655 g/day. The equation and the
  1.065 g/day statement agree with each other; 48.5% appears to be a rounding or
  transcription slip in the publication. Values in the model file are taken from
  Table 3 and left unaltered.
- **Signs recovered from Table 3, not from the printed equations.** The publisher
  PDF drops minus signs in the display equations, and the preprocessor trim
  renders every equation as `formula-not-decoded`. The Eq. 10e exponent is
  negative (-0.608 per Table 3) and the drug polynomial is `+a x^2 - b x`; both
  readings are confirmed by the paper's own arithmetic in the anchor sections
  above. The Eq. 10d indicator is `I(DAY >= 1)`; the `>=` subscript is dropped by
  every text extractor and was confirmed by rendering the page as an image.
  Reading it as "day 1 only" would switch the time effect off for the entire
  treatment period.
- **Box-Cox transform applies to the adult cohorts only.** "The Box-Cox
  transformation in the random effects for lambda was not required for the
  pediatric data. Therefore, the shape parameter was not estimated" (Results,
  Time Effects), and Table 3 labels the row "SHAPE - Adults". `model()` gates the
  transform on the complement of the pediatric-study indicators. Note this makes
  the *marginal* mean of the adult baseline rate differ from `exp(lrbase)`; the
  typical-value quantity validated above is the median, which is unaffected.
- **Study-dependent IIV standard deviation.** Table 3 reports `LN v` as the
  natural log of the standard deviation, with per-study shifts on the baseline
  rate's `LN v`. nlmixr2's `ini()` cannot express a covariate-dependent OMEGA, so
  `etalrbase` is declared standard normal with `~ fix(1)` and rescaled inside
  `model()` by `exp(lsd_rbase + ...)`. The realised SDs are 0.727 (adults), 1.306
  (Study 118), 1.763 (Study 192) and 1.040 (Study 221).
- **Time origin.** `t` is days relative to the first randomized dose; baseline
  run-in records carry `t < 1`. There are no dosing events -- `DOSE_VGB_MGD` is a
  per-record covariate, not an `amt`.
- **Study 118 dosing-weight cap.** Participants over 60 kg received their mg/kg
  dosage as if they weighed 60 kg. This belongs in the dataset that fills
  `DOSE_VGB_MGD`, not in `model()`. The 44-kg median child used in the Table 4
  reproduction is below the cap, so it does not bind here.
- **Covariates screened but not retained.** Creatinine clearance (on normalized
  dosage) and eight concomitant-AED indicators (on baseline rate) were tested and
  dropped; no point estimates are published, so they cannot be reconstructed.
  They are recorded in `covariatesDataExcluded` rather than `covariateData`.
- **Table 4 covariates and durations are assumptions.** The paper's posterior
  predictive check conditioned on the actual per-subject dosing and covariate
  records, which are not published. The reproduction above substitutes the
  paper's own median demographics (Table 2) and the Table 1 design durations. The
  sensitivity table shows all 26 gated cells stay inside the published intervals
  across the full plausible duration range, so the conclusion does not rest on
  the midpoint choice.
- **`seizure_count` observation name.** Registered as a canonical PD output in
  `inst/references/compartment-names.md`, a member of the existing per-interval
  event-count family (`hae_attacks`, `cel_count`, `migraineDays`,
  `msHeadacheDays`, `score`). `Cc` is reserved for drug-concentration outputs and
  would be wrong here.
