---
title: "Alirocumab, PCSK9 and LDL-C TMDD-QSS PK/PD (Nolain 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alirocumab, PCSK9 and LDL-C TMDD-QSS PK/PD (Nolain 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Nolain_2022_alirocumab"))$reference`
* Article: [Eur J Drug Metab Pharmacokinet 47:789-802 (2022)](https://doi.org/10.1007/s13318-022-00787-4) (open access)

Nolain et al. 2022 fitted, in one step, a joint model of total alirocumab,
total PCSK9 and LDL cholesterol (LDL-C) to nine phase I-III studies. The
pharmacokinetic layer is a two-compartment quasi-steady-state (QSS)
target-mediated drug disposition (TMDD) model in which alirocumab binds its
circulating target PCSK9; the pharmacodynamic layer is an indirect-response
model in which **free PCSK9** (not the drug) inhibits LDL-C degradation
through a steep sigmoid Imax function. Removing free PCSK9 therefore speeds up
LDL-C clearance from its pre-treatment rate `kout(0)` towards `kout`.

```{r model}
mod <- readModelDb("Nolain_2022_alirocumab")
ui <- rxode2::rxode2(mod)
ui
```

# Population

The modelling dataset held 527 alirocumab-treated subjects (13,731
observations) from nine studies (Table 1): healthy volunteers (28.5%) and
patients with primary hypercholesterolaemia (71.5%; 30.0% heterozygous
familial hypercholesterolaemia), all with LDL-C >= 100 mg/dL. Mean (SD)
characteristics (Table 2): age 52.5 (13.0) years, body weight 80.6 (16.4) kg,
46.9% female, baseline total PCSK9 7.66 (3.06) nM (median 6.99 nM), baseline
LDL-C 140 (33.1) mg/dL. 60.9% were on a statin, 12.9% on ezetimibe and 4.7%
on a fibrate. Alirocumab was given SC at 50-300 mg (single dose, Q2W or Q4W)
and, in one phase I study, IV at 0.3-12 mg/kg.

The same information is available as `ui$population`.

# Source trace

Every `ini()` value carries an in-file comment naming its location in the
paper. The equations follow Table 3.

| Element | Value | Source |
|---|---|---|
| CL | 0.221 L/day | Table 4 |
| Vc (no statin) | 3.20 L | Table 4 |
| Q | 0.557 L/day | Table 4 |
| Vp | 2.61 L (fixed) | Table 4, Sect. 3.2.1 |
| ka | 0.346 /day | Table 4 |
| F | 68.1% (logit-normal) | Table 4, Sect. 3.2.1 |
| LAG | 0.0283 day | Table 4 |
| kclear (`kint`) | 0.127 /day | Table 4 |
| kdeg | 1.34 /day | Table 4 |
| kD (`kd`) | 0.58 nM (fixed, in vitro) | Table 4, Table 3 |
| kon (`k1`) | 559 /nM/day (fixed) | Table 4, Sect. 3.2.1 |
| kout | 0.260 /day | Table 4 |
| Imax (no statin) | 74.1% (logit-normal) | Table 4 |
| IC50 of free PCSK9 (`ki50`) | 6.03 nM | Table 4 |
| Hill coefficient gamma | 11.6 | Table 4 |
| Statin on Vc | x 1.75 | Table 4, Eq. 3 |
| Statin on Imax | + 0.140 | Table 4, Eq. 5 |
| Baseline total PCSK9 on IC50 | (TBSPCSK9 / 6.99)^0.930 | Table 4, Eq. 6 |
| IIV variances (CL, kclear, kdeg, Vc, ka, F, kout, Imax, IC50) | 0.270, 0.0554, 0.124, 0.0648, 0.344, 0.626, 0.256, 0.146, 0.00578 | Table 4 |
| Residual error (add / prop) | alirocumab 0.426 nM / 25.5%; PCSK9 1.07 nM / 27.9%; LDL-C 5.71 mg/dL / 14.2% | Table 4 |
| Depot, total alirocumab, peripheral ODEs | -- | Table 3 |
| kss = kD + kclear / kon; QSS free drug | -- | Table 3 |
| Total PCSK9 ODE, ksyn = kdeg * [Ptot]baseline, Ptot(0) = [Ptot]baseline * Vc | -- | Table 3 |
| LDL-C ODE, kout(0), kin = kout(0) * [LDLC]baseline | -- | Table 3 |

# Virtual cohort

Covariates are drawn to match Table 2: 61% statin users, baseline total
PCSK9 log-normal with mean 7.66 nM and SD 3.06 nM truncated to the observed
2.36-19.6 nM, and baseline LDL-C log-normal with mean 140 mg/dL and SD
33.1 mg/dL truncated to 88.5-356 mg/dL. The three regimens are the ones the
paper simulates in Figure 4: 75 mg SC Q2W, 150 mg SC Q2W and 300 mg SC Q4W,
each for 12 weeks.

```{r cohort}
rxode2::rxSetSeed(20221016)
n_per_arm <- 150

rlnorm_ms <- function(n, mean, sd, lo, hi) {
  sdlog <- sqrt(log(1 + (sd / mean)^2))
  x <- rlnorm(n, log(mean) - sdlog^2 / 2, sdlog)
  pmin(pmax(x, lo), hi)
}

regimens <- tibble::tribble(
  ~treatment,    ~amt, ~ii, ~ndose,
  "75 mg Q2W",     75,  14,      6,
  "150 mg Q2W",   150,  14,      6,
  "300 mg Q4W",   300,  28,      3
)

cov <- tibble(id = seq_len(n_per_arm * nrow(regimens))) |>
  mutate(
    treatment = rep(regimens$treatment, each = n_per_arm),
    CONMED_STATIN = rbinom(n(), 1, 0.609),
    TPCSK9_BASE = rlnorm_ms(n(), 7.66, 3.06, 2.36, 19.6),
    LDLC = rlnorm_ms(n(), 140, 33.1, 88.5, 356)
  )

obs_times <- sort(unique(c(seq(0, 28, by = 0.5), seq(29, 150, by = 1))))

make_events <- function(cov, regimens, obs_times) {
  doses <- cov |>
    inner_join(regimens, by = "treatment") |>
    group_by(id) |>
    reframe(
      treatment = treatment, CONMED_STATIN = CONMED_STATIN,
      TPCSK9_BASE = TPCSK9_BASE, LDLC = LDLC,
      time = seq(0, by = ii[1], length.out = ndose[1]), amt = amt[1]
    ) |>
    mutate(evid = 1L, cmt = "depot", dvid = NA_integer_)
  obs <- tidyr::crossing(cov, time = obs_times) |>
    mutate(amt = 0, evid = 0L, cmt = "central", dvid = 1L)
  bind_rows(doses, obs) |>
    arrange(id, time, desc(evid)) |>
    select(id, time, evid, amt, cmt, dvid, treatment, CONMED_STATIN, TPCSK9_BASE, LDLC)
}

events <- make_events(cov, regimens, obs_times)
```

# Simulation

Observation rows sit on the `central` state with `dvid = 1`; rxode2 returns
every model output (total alirocumab `Cc`, total PCSK9 `Ctotal_target`,
`ldl`, free PCSK9 `pfree`) on each of those rows. Concentrations below are
individual predictions (between-subject variability, no residual error).

```{r simulate}
sim <- rxode2::rxSolve(ui, events, sigma = NA, keep = "treatment",
                       returnType = "data.frame") |>
  mutate(
    treatment = factor(treatment, levels = regimens$treatment),
    ldl_pct = 100 * (ldl / LDLC - 1)
  )
stopifnot(!anyNA(sim$Cc), !anyNA(sim$ldl))
```

```{r vpc, fig.height = 8}
vpc <- sim |>
  select(treatment, time, `Total alirocumab (nM)` = Cc,
         `Total PCSK9 (nM)` = Ctotal_target, `Free PCSK9 (nM)` = pfree,
         `LDL-C change from baseline (%)` = ldl_pct) |>
  pivot_longer(-c(treatment, time), names_to = "output") |>
  group_by(treatment, time, output) |>
  summarise(q05 = quantile(value, 0.05), q50 = median(value),
            q95 = quantile(value, 0.95), .groups = "drop")

ggplot(vpc, aes(time, q50, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.15, colour = NA) +
  geom_line() +
  facet_wrap(~output, ncol = 1, scales = "free_y") +
  labs(x = "Time (day)", y = NULL, colour = NULL, fill = NULL,
       caption = "Median and 90% prediction interval, 150 virtual subjects per regimen.") +
  theme_bw() + theme(legend.position = "bottom")
```

# Replicate published figures

## Figure 4: typical profiles by regimen

Figure 4 shows typical alirocumab, free PCSK9 and LDL-C profiles for the three
regimens with a baseline free PCSK9 of 5 nM (read off Figure 4b). The paper
does not state whether these curves are for a statin user, so both covariate
states are drawn here.

```{r fig4, fig.height = 8}
typ_cov <- tidyr::crossing(treatment = regimens$treatment, CONMED_STATIN = 0:1) |>
  mutate(id = row_number(), TPCSK9_BASE = 5, LDLC = 140)
typ_events <- make_events(typ_cov, regimens, obs_times)
typ <- rxode2::rxSolve(ui, typ_events, omega = NA, sigma = NA,
                       keep = "treatment", returnType = "data.frame") |>
  mutate(
    treatment = factor(treatment, levels = regimens$treatment),
    statin = ifelse(CONMED_STATIN == 1, "with statin (Vc 5.6 L, Imax 88.1%)",
                    "without statin (Vc 3.2 L, Imax 74.1%)"),
    ldl_pct = 100 * (ldl / LDLC - 1)
  )

typ |>
  select(treatment, statin, time, `a: Total alirocumab (nM)` = Cc,
         `b: Free PCSK9 (nM)` = pfree, `c: LDL-C change from baseline (%)` = ldl_pct) |>
  pivot_longer(-c(treatment, statin, time), names_to = "panel") |>
  ggplot(aes(time, value, colour = treatment, linetype = statin)) +
  geom_line() +
  facet_wrap(~panel, ncol = 1, scales = "free_y") +
  labs(x = "Time (day)", y = NULL, colour = NULL, linetype = NULL,
       caption = "Replicates Figure 4 of Nolain 2022 (typical values, baseline PCSK9 5 nM).") +
  theme_bw() + theme(legend.position = "bottom", legend.box = "vertical")
```

The values digitised from Figure 4 fall between the two typical curves:

```{r fig4-check}
fig4 <- typ |>
  group_by(treatment, CONMED_STATIN) |>
  summarise(
    cmax_first = max(Cc[time < 14]),
    pfree_day14 = pfree[time == 13.5],
    ldl_plateau = min(ldl_pct),
    .groups = "drop"
  )

fig4_75 <- fig4 |> filter(treatment == "75 mg Q2W") |> arrange(CONMED_STATIN)
fig4_digitised <- c(cmax_first = 43, pfree_day14 = 4.1, ldl_plateau = -68.6)

fig4_75 |>
  select(CONMED_STATIN, cmax_first, pfree_day14, ldl_plateau) |>
  mutate(across(-CONMED_STATIN, ~ signif(.x, 3))) |>
  bind_rows(tibble(CONMED_STATIN = NA, !!!as.list(fig4_digitised))) |>
  mutate(Curve = c("Model, no statin", "Model, statin", "Figure 4 (digitised)")) |>
  select(Curve, `First-dose Cmax (nM)` = cmax_first,
         `Free PCSK9 before dose 2 (nM)` = pfree_day14,
         `LDL-C plateau (%)` = ldl_plateau) |>
  knitr::kable(caption = "75 mg SC Q2W, baseline PCSK9 5 nM.")

stopifnot(
  # Each digitised value lies inside the [statin, no-statin] typical envelope.
  fig4_digitised[["cmax_first"]] > fig4_75$cmax_first[2],
  fig4_digitised[["cmax_first"]] < fig4_75$cmax_first[1],
  fig4_digitised[["pfree_day14"]] > fig4_75$pfree_day14[1],
  fig4_digitised[["pfree_day14"]] < fig4_75$pfree_day14[2],
  fig4_digitised[["ldl_plateau"]] > fig4_75$ldl_plateau[2],
  fig4_digitised[["ldl_plateau"]] < fig4_75$ldl_plateau[1]
)
```

## Figure 2: size of each retained covariate effect

Figure 2 quantifies each covariate relationship by switching it on alone at
75 mg Q2W and reading alirocumab AUC over weeks 10-12 and the LDL-C decrease
at week 12, relative to a reference subject (no statin, baseline PCSK9 at the
median 6.99 nM). The paper reports an AUC decrease of about 45% for the statin
effect on Vc (Sect. 3.2.2); the LDL-C distributions for the statin effects on
Vc and on Imax are centred near -20% and +20% (Fig. 2a; 42% and 48%
probability of exceeding a 20% change). The AUC is computed with PKNCA.

```{r fig2}
fig2_scenarios <- list(
  "Reference" = list(mod = ui, statin = 0, ic50_pcsk9 = 6.99),
  "Statin on Vc only" = list(mod = rxode2::ini(ui, e_conmed_statin_imax = 0), statin = 1, ic50_pcsk9 = 6.99),
  "Statin on Imax only" = list(mod = rxode2::ini(ui, e_conmed_statin_vc = 1), statin = 1, ic50_pcsk9 = 6.99),
  "IC50 at baseline PCSK9 9.10 nM (p75) only" = list(mod = ui, statin = 0, ic50_pcsk9 = 9.10)
)

fig2_cov <- tibble(id = 1L, treatment = "75 mg Q2W", CONMED_STATIN = 0L,
                   TPCSK9_BASE = 6.99, LDLC = 140)
fig2_events <- make_events(fig2_cov, regimens, seq(0, 84, by = 0.25))

fig2_sim <- bind_rows(lapply(names(fig2_scenarios), function(nm) {
  sc <- fig2_scenarios[[nm]]
  # The p75 scenario moves only the covariate inside the IC50 equation (Eq. 6)
  # and keeps the PCSK9 turnover at the median baseline, as Figure 2b does.
  m <- rxode2::ini(sc$mod, lki50 = log(6.03 * (sc$ic50_pcsk9 / 6.99)^0.930))
  ev <- mutate(fig2_events, CONMED_STATIN = sc$statin)
  rxode2::rxSolve(m, ev, omega = NA, sigma = NA, returnType = "data.frame") |>
    mutate(scenario = nm, id = 1L)
}))

conc_obj <- PKNCA::PKNCAconc(
  fig2_sim |> filter(!is.na(Cc)) |> select(scenario, id, time, Cc),
  Cc ~ time | scenario + id
)
dose_obj <- PKNCA::PKNCAdose(
  fig2_events |> filter(evid == 1) |> select(id, time, amt) |>
    tidyr::crossing(scenario = names(fig2_scenarios)),
  amt ~ time | scenario + id
)
fig2_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  conc_obj, dose_obj,
  intervals = data.frame(start = 70, end = 84, auclast = TRUE, cmax = TRUE)
))

fig2_tab <- as.data.frame(fig2_nca$result) |>
  filter(PPTESTCD == "auclast") |>
  select(scenario, auc = PPORRES) |>
  left_join(
    fig2_sim |> filter(time == 84) |> transmute(scenario, ldl_drop = 100 * (1 - ldl / 140)),
    by = "scenario"
  ) |>
  mutate(
    auc_change = 100 * (auc / auc[scenario == "Reference"] - 1),
    ldl_change = 100 * (ldl_drop / ldl_drop[scenario == "Reference"] - 1)
  ) |>
  arrange(factor(scenario, levels = names(fig2_scenarios)))

fig2_tab |>
  mutate(across(where(is.numeric), ~ round(.x, 1))) |>
  select(Scenario = scenario, `AUC wk 10-12 (nM*day)` = auc,
         `AUC change (%)` = auc_change, `LDL-C decrease wk 12 (%)` = ldl_drop,
         `Change in LDL-C decrease (%)` = ldl_change) |>
  knitr::kable()

get_row <- function(nm) fig2_tab[fig2_tab$scenario == nm, ]
stopifnot(
  # Statin on Vc: paper reports approximately -45% AUC.
  abs(get_row("Statin on Vc only")$auc_change - (-45)) < 5,
  # Statin effects on LDL-C decrease: Figure 2a centres near -20% and +20%.
  abs(get_row("Statin on Vc only")$ldl_change - (-20)) < 5,
  abs(get_row("Statin on Imax only")$ldl_change - 20) < 5,
  # Imax does not enter the PK, so its AUC equals the reference up to the
  # ODE solver tolerance.
  abs(get_row("Statin on Imax only")$auc_change) < 0.01
)
```

For the IC50 relationship the paper states that subjects with a baseline
PCSK9 at or above 9.10 nM would have a lipid-lowering effect reduced by about
60% versus the median subject (Sect. 3.2.2); Figure 2b shows a wide
distribution centred near -60%. Moving only the IC50 covariate gives a
reduction of `r round(-get_row("IC50 at baseline PCSK9 9.10 nM (p75) only")$ldl_change)`%
with the typical parameters, inside the spread of Figure 2b (which reflects
the uncertainty of the estimates, and in particular of the steep Hill
coefficient).

Note that this is a property of the single relationship, not of a real
high-PCSK9 subject. In the full model a higher baseline PCSK9 also raises the
total-PCSK9 initial condition and `kout(0)`. Because the IC50 exponent (0.930)
is close to 1, the ratio of baseline PCSK9 to IC50 hardly changes, and the
week-12 LDL-C reduction at 9.10 nM is only a few percent smaller than at the
median:

```{r p75-full}
p75_full <- rxode2::rxSolve(ui, mutate(fig2_events, TPCSK9_BASE = 9.10),
                            omega = NA, sigma = NA, returnType = "data.frame")
ref_full <- rxode2::rxSolve(ui, fig2_events, omega = NA, sigma = NA, returnType = "data.frame")
round(100 * (1 - c(median_6.99 = ref_full$ldl[ref_full$time == 84],
                   p75_9.10 = p75_full$ldl[p75_full$time == 84]) / 140), 1)
```

# Mechanistic checks

Without drug the system must stay at its pre-treatment steady state (total
PCSK9 at baseline, LDL-C at baseline). With free PCSK9 fully suppressed, the
LDL-C turnover settles at `kin / kout`, so the maximal fractional reduction is
`Imax * H0` with `H0 = P0^gamma / (IC50^gamma + P0^gamma)` evaluated at the
baseline PCSK9 `P0`. At the median baseline (6.99 nM) that is 62.8% without a
statin and 74.6% with one, consistent with the "up to 62.7%" LDL-C reduction
the paper cites from the clinical programme.

```{r mechanistic}
mech_cov <- tidyr::crossing(treatment = c("none", "300 mg Q2W"), CONMED_STATIN = 0:1) |>
  mutate(id = row_number(), TPCSK9_BASE = 6.99, LDLC = 140)
mech_regimens <- tibble(treatment = c("none", "300 mg Q2W"), amt = c(0, 300), ii = 14, ndose = c(1, 12))
mech <- rxode2::rxSolve(ui, make_events(mech_cov, mech_regimens, seq(0, 168, by = 1)),
                        omega = NA, sigma = NA, keep = "treatment", returnType = "data.frame")

imax_typ <- c(0.741, 0.741 + 0.140)
h0 <- 1 / (1 + (6.03 / 6.99)^11.6)
closed_form <- 100 * imax_typ * h0

mech_summary <- mech |>
  group_by(treatment, CONMED_STATIN) |>
  summarise(
    ptot_range = diff(range(Ctotal_target)),
    ldl_range = diff(range(ldl)),
    max_ldl_drop = 100 * (1 - min(ldl) / 140),
    .groups = "drop"
  )
mech_summary |> mutate(closed_form = closed_form[CONMED_STATIN + 1]) |> knitr::kable(digits = 3)

untreated <- filter(mech_summary, treatment == "none")
treated <- filter(mech_summary, treatment == "300 mg Q2W") |> arrange(CONMED_STATIN)
stopifnot(
  # The same drawn parameters on both sides: pure numerical error, so tight.
  all(untreated$ptot_range < 1e-6),
  all(untreated$ldl_range < 1e-6),
  all(abs(treated$max_ldl_drop - closed_form) < 0.5)
)
```

# PKNCA validation

The paper reports no non-compartmental summaries. PKNCA is run on total
alirocumab for the first dosing interval and the last (twelfth week) Q2W
interval of the virtual cohort, grouped by regimen.

```{r pknca}
nca_conc <- sim |>
  filter(!is.na(Cc)) |>
  select(id, treatment, time, Cc)
nca_dose <- events |>
  filter(evid == 1) |>
  select(id, treatment, time, amt)

intervals <- bind_rows(
  data.frame(treatment = c("75 mg Q2W", "150 mg Q2W"), start = 0, end = 14),
  data.frame(treatment = "300 mg Q4W", start = 0, end = 28),
  data.frame(treatment = c("75 mg Q2W", "150 mg Q2W"), start = 70, end = 84),
  data.frame(treatment = "300 mg Q4W", start = 56, end = 84)
) |>
  mutate(cmax = TRUE, tmax = TRUE, auclast = TRUE, cmin = TRUE)

nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(nca_conc, Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(nca_dose, amt ~ time | treatment + id),
  intervals = intervals
))

as.data.frame(nca$result) |>
  group_by(treatment, start, end, PPTESTCD) |>
  summarise(median = signif(median(PPORRES), 3), .groups = "drop") |>
  pivot_wider(names_from = PPTESTCD, values_from = median) |>
  select(Regimen = treatment, Start = start, End = end,
         `Cmax (nM)` = cmax, `Tmax (day)` = tmax,
         `Cmin (nM)` = cmin, `AUClast (nM*day)` = auclast) |>
  knitr::kable(caption = "Median NCA of total alirocumab in the virtual cohort.")
```

# Assumptions and deviations

* **Molecular weight.** Nolain 2022 doses in nmol and does not state the
  alirocumab molecular weight. The model accepts doses in mg and converts with
  146,000 g/mol, the approximate molecular weight on the Praluent label (the
  value also used by `Sokolov_2019_antipcsk9_qsp`). `Djebli_2017_alirocumab`
  uses 144,100 g/mol; the 1.3% difference is negligible against the 68.1%
  bioavailability.
* **IV dosing.** IV doses go to `central` in mg and are converted to nmol
  with the same factor (bioavailability 1). The paper's IV infusion term
  `In(t)` is reproduced by an infusion record on `central`.
* **Statin effect on Imax.** Eq. 5 adds the statin coefficient to the
  typical Imax (74.1% + 14.0% = 88.1%, matching the 88% in the abstract and
  Sect. 3.2.3). The logit-normal between-subject variability is applied
  around that sum.
* **Free-drug QSS root.** Table 3's free alirocumab expression is evaluated
  in the algebraically identical form `2 kss [Atot] / (sqrt(.) - a)` when
  `a = [Atot] - [Ptot] - kss` is negative, which avoids cancellation when the
  drug is far below the target.
* **Parameter names.** The paper's complex clearance rate `kclear` is stored
  as the canonical `kint`, `kon` as `k1`, `kD` as `kd`, and the IC50 of
  **free PCSK9** (a target concentration, not a drug concentration) as `ki50`.
* **Dropped covariates.** The statin effect on kout, the sex effect on Imax
  and the baseline-PCSK9 effect on kdeg were tested a priori and removed in
  the covariate reduction step (Sect. 3.2.2); they are not encoded.
* **Figure 4 covariates.** Figure 4 does not state the statin status of the
  simulated subject. Its curves fall between the no-statin and statin typical
  curves for every panel read (first-dose peak, free PCSK9 before dose 2,
  LDL-C plateau), and a population median with the dataset's 61% statin mix
  does not reproduce them either. The figure is therefore compared as an
  envelope, not point by point.
* **Supplement.** The supplement (base-model estimates, VPCs) was not used;
  the final model is fully specified by Table 3, Eqs. 3-7 and Table 4.
* **Virtual cohort.** Baseline PCSK9 and LDL-C are drawn independently from
  log-normal distributions matched to the Table 2 means and SDs; the
  correlation between them is not reported.
