---
title: "Doripenem (Nonoshita 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Doripenem (Nonoshita 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Nonoshita_2020_doripenem"))$reference`
* Description: `r rxode2::rxode(readModelDb("Nonoshita_2020_doripenem"))$description`
* Article: [Sci Rep. 2020;10:22148](https://doi.org/10.1038/s41598-020-79076-6) (open access)

**Read the "Assumptions and deviations" section before using this model.** The
parameter values below are transcribed exactly as the paper prints them, but
they predict roughly twice the plasma concentrations that the paper's own
visual predictive check and goodness-of-fit plots show for the same 500 mg
dose. The comparison is in the "Visual predictive check" section.

# Population

Twenty-one adult inpatients in the intensive care unit of Oita University
Hospital, Japan, treated with doripenem for severe infection (bacteremia,
sepsis or septic shock, infective endocarditis, pneumonia, intraperitoneal or
urinary-tract infection, postoperative infection). There were 18 men and 3
women, age 61.8 +/- 18.9 years (range 15-86), body weight 61.5 +/- 13.6 kg
(range 28.8-92.4), APACHE II 17.6 +/- 6.7 and SOFA 7.5 +/- 2.6 (Table 1 and
Supplementary Table S1). Cockcroft-Gault creatinine clearance was 68.0 +/- 33.4
mL/min (range 20.7-155.6).

Nine patients were on continuous renal replacement therapy (CRRT), given as
continuous hemodiafiltration through a cellulose triacetate membrane with a
blood flow of 80-100 mL/min, dialysate and replacement-fluid flows of 0.3-0.9
L/h each, and a filtrate flow (QE) of 0.6-1.8 L/h. Creatinine clearance was
57.3 +/- 26.5 mL/min in the CRRT stratum and 76.0 +/- 35.8 mL/min in the other
12 patients.

Twenty patients received 500 mg and one received 250 mg as a 1 h infusion.
Plasma was sampled before and 1, 2, 4, 6 and 8 h after the start of the first
infusion only (97 samples).

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Nonoshita_2020_doripenem")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Nonoshita_2020_doripenem.R`. The table
below collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Two-compartment model, IV infusion, `CL_body` and `CL_CRRT` both from the central compartment | n/a | Methods "Population pharmacokinetics"; Figure 1 |
| `CL_total = CL_body(non-CRRT) = 3.65 x (Ccr/62.25)^0.64` without CRRT | n/a | Abstract; Results "final model" paragraph |
| `CL_total = CL_body(CRRT) + CL_CRRT = 2.49 x (Ccr/52.75)^0.42 + CL_CRRT` with CRRT | n/a | Abstract; Results "final model" paragraph |
| `CL_CRRT = QE x 0.919` | n/a | Abstract; Results "final model" paragraph |
| `lcl_offcrrt` | 3.65 L/h | Table 3, population mean |
| `lcl_oncrrt` | 2.49 L/h | Table 3, population mean |
| `lvc` | 10.04 L | Table 3, population mean |
| `lvp` | 8.13 L | Table 3, population mean |
| `lq` | 3.53 L/h | Table 3, population mean |
| `e_crcl_cl_offcrrt` | 0.64 | Results final-model equation |
| `e_crcl_cl_oncrrt` | 0.42 | Results final-model equation |
| `fu` | 0.919 (fixed, from the literature) | Results final-model equation (Hori 2006) |
| CRCL centring values | 62.25 and 52.75 mL/min (stratum medians) | Results final-model equation; Discussion |
| `etalcl_offcrrt` | 7.3% CV -> 0.0053149 | Table 3; Discussion |
| `etalcl_oncrrt` | 22.2% CV -> 0.048108 | Table 3; Discussion |
| `etalvc` | 13.2% CV -> 0.017274 | Table 3; Results |
| `etalvp` | 24.2% CV -> 0.056913 | Table 3; Results |
| `etalq` | 12.7% CV -> 0.016000 | Table 3; Results |
| `addSd` | 0.70 mg/L | Table 3, residual variability estimate |
| `propSd` | 0.365 | Table 3; Results "residual variability was 36.5%" |

Inter-individual variability was modelled exponentially (Methods), so each
percentage is converted with `omega^2 = log(CV^2 + 1)`.

# Typical-value checks

The model is deterministic in the typical subject, so the paper's clearance
arithmetic can be checked exactly. The Discussion reports that the nine CRRT
patients had a total clearance of 3.81 L/h and a CRRT clearance of 1.31 L/h
(means of the individual estimates). At the CRRT-stratum median creatinine
clearance of 52.75 mL/min the body clearance is the typical 2.49 L/h, and a CRRT
clearance of 1.31 L/h corresponds to a filtrate flow of
`1.31 / 0.919 = 1.43` L/h, inside the 0.6-1.8 L/h range the Methods report.

```{r typical}
mod <- rxode2::rxode2(readModelDb("Nonoshita_2020_doripenem"))
mod_typ <- rxode2::zeroRe(mod)

typ_cov <- tibble::tribble(
  ~scenario,                                ~RRT_CRRT_STATUS, ~CRCL, ~RRT_CRRT_EFFLUENT_FLOW,
  "No CRRT, CRCL 62.25 mL/min",                           0,  62.25,                       0,
  "CRRT, CRCL 52.75 mL/min, QE 0.6 L/h",                  1,  52.75,                     600,
  "CRRT, CRCL 52.75 mL/min, QE 1.43 L/h",                 1,  52.75,           1310 / 0.919,
  "CRRT, CRCL 52.75 mL/min, QE 1.8 L/h",                  1,  52.75,                    1800
) |>
  dplyr::mutate(id = dplyr::row_number())

ev_typ <- rxode2::et(amt = 500, dur = 1, cmt = "central") |>
  rxode2::et(c(1, 8), cmt = "central") |>
  rxode2::et(id = typ_cov$id) |>
  as.data.frame() |>
  dplyr::left_join(typ_cov, by = "id")

sim_typ <- rxode2::rxSolve(mod_typ, events = ev_typ, returnType = "data.frame")

typ_tab <- sim_typ |>
  dplyr::filter(time == 1) |>
  dplyr::left_join(typ_cov |> dplyr::select(id, scenario), by = "id") |>
  dplyr::mutate(crrt_share = 100 * cl_crrt / cl) |>
  dplyr::select(scenario, cl_body, cl_crrt, cl, crrt_share)

typ_tab |>
  dplyr::rename(
    "Scenario" = scenario,
    "CL body (L/h)" = cl_body,
    "CL CRRT (L/h)" = cl_crrt,
    "CL total (L/h)" = cl,
    "CRRT share of CL total (%)" = crrt_share
  ) |>
  knitr::kable(digits = c(0, 3, 3, 3, 1))

typ_val <- function(col, which) {
  v <- typ_tab[[col]][typ_tab$scenario == which]
  if (length(v) != 1L) stop("no unique row for '", which, "'")
  v
}
stopifnot(
  # The printed typical values are returned at each stratum's centring value.
  abs(typ_val("cl", "No CRRT, CRCL 62.25 mL/min") - 3.65) < 1e-6,
  abs(typ_val("cl_body", "CRRT, CRCL 52.75 mL/min, QE 1.43 L/h") - 2.49) < 1e-6,
  # CL_CRRT = QE x 0.919 and CL_total = CL_body + CL_CRRT reproduce the
  # Discussion's 1.31 and 3.81 L/h (the latter to its printed precision).
  abs(typ_val("cl_crrt", "CRRT, CRCL 52.75 mL/min, QE 1.43 L/h") - 1.31) < 1e-6,
  abs(typ_val("cl", "CRRT, CRCL 52.75 mL/min, QE 1.43 L/h") - 3.81) < 0.02
)
```

Across the reported 0.6-1.8 L/h filtrate range the CRRT arm contributes 18-40%
of total clearance for a typical CRRT patient. The paper summarises this as
30-40% (Abstract) and 40.0 +/- 3.92% (mean +/- SE, Discussion); the latter is a
mean of per-patient ratios, which need not equal the ratio of the typical
values.

# Virtual cohort

Individual data are not published. The cohort below reproduces the study
design: a single 500 mg dose as a 1 h infusion, with 12 non-CRRT and 9 CRRT
patients scaled to 200 and 150 virtual subjects (the 12:9 ratio of Table 1).
Creatinine clearance is drawn per stratum from a log-normal distribution
parameterised with the Table 1 mean and SD, restricted to the observed
20.7-155.6 mL/min range by redrawing; the truncation trims the upper tail, so
the realised non-CRRT mean is a little below the 76.0 mL/min of Table 1. The
filtrate flow of CRRT subjects is drawn uniformly over
the reported 0.6-1.8 L/h.

```{r cohort}
# rxode2's simulation RNG is partitioned per solver thread, so the drawn
# cohort is not reproducible across machines with different thread counts.
# Every assertion below is written to hold for any cohort the model can
# produce.
rxode2::rxSetSeed(20201218)
set.seed(20201218)

draw_crcl <- function(n, mean, sd, lo = 20.7, hi = 155.6) {
  s2 <- log(1 + (sd / mean)^2)
  mu <- log(mean) - s2 / 2
  out <- numeric(0)
  while (length(out) < n) {
    x <- stats::rlnorm(2 * n, mu, sqrt(s2))
    out <- c(out, x[x >= lo & x <= hi])
  }
  out[seq_len(n)]
}

N_OFF <- 200L
N_ON <- 150L
cohort <- dplyr::bind_rows(
  tibble::tibble(
    id = seq_len(N_OFF),
    arm = "No CRRT",
    RRT_CRRT_STATUS = 0,
    CRCL = draw_crcl(N_OFF, 76.0, 35.8),
    RRT_CRRT_EFFLUENT_FLOW = 0
  ),
  tibble::tibble(
    id = N_OFF + seq_len(N_ON),
    arm = "CRRT",
    RRT_CRRT_STATUS = 1,
    CRCL = draw_crcl(N_ON, 57.3, 26.5),
    RRT_CRRT_EFFLUENT_FLOW = stats::runif(N_ON, 600, 1800)
  )
)

cohort |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    n = dplyr::n(),
    crcl_mean = mean(CRCL),
    crcl_sd = stats::sd(CRCL),
    qe_mean_L_h = mean(RRT_CRRT_EFFLUENT_FLOW) / 1000
  ) |>
  dplyr::rename(
    "Arm" = arm,
    "N" = n,
    "CRCL mean (mL/min)" = crcl_mean,
    "CRCL SD (mL/min)" = crcl_sd,
    "QE mean (L/h)" = qe_mean_L_h
  ) |>
  knitr::kable(digits = 1)
```

# Visual predictive check (Figure 5)

The solid line of Figure 5 is the median of the observed concentrations at each
nominal time. The values below were digitised by the maintainers from the
published figure and are approximate (read to about 0.5 mg/L).

```{r vpc}
fig5_obs_median <- tibble::tibble(
  time = c(1, 2, 4, 6, 8),
  obs_median = c(21.0, 10.8, 4.9, 3.0, 2.1)
)

ev_vpc <- rxode2::et(amt = 500, dur = 1, cmt = "central") |>
  rxode2::et(seq(0, 8, by = 0.25), cmt = "central") |>
  rxode2::et(id = cohort$id) |>
  as.data.frame() |>
  dplyr::left_join(cohort, by = "id")

sim_vpc <- rxode2::rxSolve(mod, events = ev_vpc, returnType = "data.frame", keep = "arm")

vpc_sum <- sim_vpc |>
  dplyr::filter(time > 0) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    p05 = stats::quantile(sim, 0.05),
    p50 = stats::quantile(sim, 0.50),
    p95 = stats::quantile(sim, 0.95),
    .groups = "drop"
  )

ggplot(vpc_sum, aes(time)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "grey85") +
  geom_line(aes(y = p50), colour = "grey30") +
  geom_point(
    data = fig5_obs_median, aes(y = obs_median),
    shape = 21, size = 2.5, fill = "white"
  ) +
  geom_line(data = fig5_obs_median, aes(y = obs_median), linetype = "dashed") +
  labs(
    x = "Time after start of infusion (h)",
    y = "Doripenem plasma concentration (mg/L)",
    title = "Replicates Figure 5 of Nonoshita 2020",
    caption = paste(
      "Grey band and line: simulated 5th-95th percentiles and median (with residual error).",
      "Points and dashed line: observed median digitised from Figure 5.",
      sep = "\n"
    )
  ) +
  theme_bw()

vpc_cmp <- vpc_sum |>
  dplyr::inner_join(fig5_obs_median, by = "time") |>
  dplyr::mutate(ratio = p50 / obs_median)
stopifnot(nrow(vpc_cmp) == 5L)

vpc_cmp |>
  dplyr::select(time, obs_median, p50, ratio) |>
  dplyr::rename(
    "Time (h)" = time,
    "Figure 5 observed median (mg/L)" = obs_median,
    "Simulated median (mg/L)" = p50,
    "Simulated / observed" = ratio
  ) |>
  knitr::kable(digits = 2)
```

The simulated median sits about 1.7-2.3-fold above the observed median at every
sampling time. This is not a property of the virtual cohort: the typical
non-CRRT subject alone (V1 = 10.04 L, CL = 3.65 L/h) reaches
`r signif(sim_typ$Cc[sim_typ$id == 1 & sim_typ$time == 1], 3)` mg/L at the end
of the 500 mg infusion, while the paper's own population predictions (Figure 4,
upper left) do not exceed about 30 mg/L and cluster around 20 mg/L at 1 h. The
reasons are set out under "Assumptions and deviations"; the parameters have not
been adjusted.

```{r vpc-gate}
# The discrepancy is structural (every time point, same direction), so the
# centre of the ratios is a stable property of the printed parameters. A
# change to any transcribed clearance or volume moves it by tens of percent.
stopifnot(
  stats::median(vpc_cmp$ratio) > 1.4,
  stats::median(vpc_cmp$ratio) < 3.0
)
```

# Non-compartmental analysis

The paper reports no NCA. PKNCA is used here as an internal consistency check:
after a single dose, `Dose / AUC0-inf` must return each subject's total
clearance. The simulation uses the individual predictions (no residual error)
out to 48 h so that the extrapolated fraction of AUC is small.

```{r nca}
ev_nca <- rxode2::et(amt = 500, dur = 1, cmt = "central") |>
  rxode2::et(c(seq(0, 2, by = 0.1), seq(2.25, 12, by = 0.25), seq(12.5, 48, by = 0.5)), cmt = "central") |>
  rxode2::et(id = cohort$id) |>
  as.data.frame() |>
  dplyr::left_join(cohort, by = "id")

sim_nca <- rxode2::rxSolve(
  mod, events = ev_nca, returnType = "data.frame", keep = "arm",
  rtol = 1e-10, atol = 1e-12
)
stopifnot(all(sim_nca$Cc >= -1e-6 * max(sim_nca$Cc, na.rm = TRUE), na.rm = TRUE))

conc_df <- sim_nca |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(Cc = pmax(Cc, 0), treatment = arm) |>
  dplyr::select(id, time, Cc, treatment)

dose_df <- cohort |>
  dplyr::transmute(id, time = 0, amt = 500, treatment = arm)

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | treatment + id, concu = "mg/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id, doseu = "mg", route = "intravascular", duration = 1)
intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE, cl.obs = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_df <- as.data.frame(nca_res)
summary(nca_res)
```

```{r nca-gate}
cl_nca <- nca_df |>
  dplyr::filter(PPTESTCD == "cl.obs") |>
  dplyr::select(id, treatment, cl_nca = PPORRES)
cl_ind <- sim_nca |>
  dplyr::filter(time == 1) |>
  dplyr::select(id, cl)
cl_chk <- dplyr::inner_join(cl_nca, cl_ind, by = "id") |>
  dplyr::mutate(pct_diff = 100 * (cl_nca / cl - 1))
stopifnot(nrow(cl_chk) == N_OFF + N_ON)

cl_chk |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    median_cl_model = stats::median(cl),
    median_cl_nca = stats::median(cl_nca),
    median_pct_diff = stats::median(pct_diff),
    .groups = "drop"
  ) |>
  dplyr::rename(
    "Arm" = treatment,
    "Median model CL (L/h)" = median_cl_model,
    "Median NCA Dose/AUCinf (L/h)" = median_cl_nca,
    "Median difference (%)" = median_pct_diff
  ) |>
  knitr::kable(digits = 2)

stopifnot(
  abs(stats::median(cl_chk$pct_diff)) < 2,
  stats::quantile(abs(cl_chk$pct_diff), 0.9) < 5
)
```

# Probability of target attainment (Table 4)

Table 4 gives the probability of attaining 40% and 100% of the dosing interval
with free doripenem above an MIC of 2 mg/L (and 100% above 4 x MIC) at steady
state, for 1 h intermittent infusion (InI) and 4 h extended infusion (ExI) every
8 h, and for continuous infusion (CI) of the same daily dose. Free
concentration is total concentration times the 0.919 unbound fraction.

The Methods do not state the virtual-cohort sizes, the creatinine-clearance
distribution within each band, the filtrate flow used for CRRT subjects, or
whether residual error was added. This replication draws creatinine clearance
uniformly within each band (the 0-30 band from 10 mL/min), uses a filtrate flow
of 1.2 L/h (the middle of the reported range) and simulates without residual
error, 200 subjects per arm.

```{r pta}
rxode2::rxSetSeed(20201219)
set.seed(20201219)
N_PTA <- 200L
MIC <- 2
FU <- 0.919

bands <- tibble::tribble(
  ~band,          ~lo, ~hi, ~crrt,
  "0 < Ccr <= 30",  10,  30,     1,
  "0 < Ccr <= 30",  10,  30,     0,
  "30 < Ccr <= 60", 30,  60,     1,
  "30 < Ccr <= 60", 30,  60,     0,
  "60 < Ccr <= 90", 60,  90,     0
)
pta_arms <- tidyr::expand_grid(bands, method = c("InI", "ExI"), dose = c(250, 500, 1000, 2000)) |>
  dplyr::mutate(arm_id = dplyr::row_number())

make_pta_arm <- function(a) {
  ids <- (a$arm_id - 1L) * N_PTA + seq_len(N_PTA)
  covs <- tibble::tibble(
    id = ids,
    CRCL = stats::runif(N_PTA, a$lo, a$hi),
    RRT_CRRT_STATUS = a$crrt,
    RRT_CRRT_EFFLUENT_FLOW = 1200 * a$crrt,
    arm_id = a$arm_id
  )
  rxode2::et(amt = a$dose, dur = if (a$method == "InI") 1 else 4, ii = 8, ss = 1, cmt = "central") |>
    rxode2::et(round(seq(0, 8, by = 0.1), 6), cmt = "central") |>
    rxode2::et(id = ids) |>
    as.data.frame() |>
    dplyr::left_join(covs, by = "id")
}
ev_pta <- dplyr::bind_rows(lapply(split(pta_arms, pta_arms$arm_id), make_pta_arm))

sim_pta <- rxode2::rxSolve(
  mod, events = ev_pta, returnType = "data.frame", keep = "arm_id",
  maxsteps = 1e6
)

pta_sim <- sim_pta |>
  dplyr::filter(time > 0) |>
  dplyr::group_by(arm_id, id) |>
  dplyr::summarise(
    ft_mic = mean(FU * Cc > MIC),
    ft_mic4 = mean(FU * Cc > 4 * MIC),
    .groups = "drop"
  ) |>
  dplyr::group_by(arm_id) |>
  dplyr::summarise(
    pta40 = 100 * mean(ft_mic >= 0.4),
    pta100 = 100 * mean(ft_mic >= 1),
    pta100x4 = 100 * mean(ft_mic4 >= 1),
    .groups = "drop"
  ) |>
  dplyr::left_join(pta_arms, by = "arm_id")

# Published Table 4, InI and ExI rows (40% fT>MIC, 100% fT>MIC, 100% fT>MICx4).
table4 <- tibble::tribble(
  ~band,            ~crrt, ~method, ~dose, ~pub40, ~pub100, ~pub100x4,
  "0 < Ccr <= 30",      1, "InI",   2000,   99.7,    86.6,     81.5,
  "0 < Ccr <= 30",      1, "InI",   1000,   99.5,    81.5,     61.6,
  "0 < Ccr <= 30",      1, "InI",    500,   99.0,    73.7,     39.9,
  "0 < Ccr <= 30",      1, "InI",    250,   98.2,    61.6,     12.2,
  "0 < Ccr <= 30",      0, "InI",   2000,  100.0,    97.5,     94.5,
  "0 < Ccr <= 30",      0, "InI",   1000,  100.0,    96.4,     91.0,
  "0 < Ccr <= 30",      0, "InI",    500,  100.0,    94.5,     82.8,
  "0 < Ccr <= 30",      0, "InI",    250,   99.0,    91.0,     66.7,
  "0 < Ccr <= 30",      1, "ExI",   2000,  100.0,   100.0,    100.0,
  "0 < Ccr <= 30",      1, "ExI",   1000,  100.0,   100.0,     99.4,
  "0 < Ccr <= 30",      1, "ExI",    500,  100.0,   100.0,     94.3,
  "0 < Ccr <= 30",      1, "ExI",    250,  100.0,    99.4,     58.8,
  "0 < Ccr <= 30",      0, "ExI",   2000,  100.0,   100.0,    100.0,
  "0 < Ccr <= 30",      0, "ExI",   1000,  100.0,   100.0,    100.0,
  "0 < Ccr <= 30",      0, "ExI",    500,  100.0,   100.0,     99.8,
  "0 < Ccr <= 30",      0, "ExI",    250,  100.0,   100.0,     91.4,
  "30 < Ccr <= 60",     1, "InI",   2000,   98.9,    74.8,     58.8,
  "30 < Ccr <= 60",     1, "InI",   1000,   98.5,    68.5,     44.4,
  "30 < Ccr <= 60",     1, "InI",    500,   97.8,    58.8,     24.1,
  "30 < Ccr <= 60",     1, "InI",    250,   95.8,    44.4,      5.1,
  "30 < Ccr <= 60",     0, "InI",   2000,   99.8,    87.3,     75.9,
  "30 < Ccr <= 60",     0, "InI",   1000,   99.6,    82.9,     63.9,
  "30 < Ccr <= 60",     0, "InI",    500,   99.4,    75.9,     44.3,
  "30 < Ccr <= 60",     0, "InI",    250,   99.0,    63.9,     20.6,
  "30 < Ccr <= 60",     1, "ExI",   2000,  100.0,   100.0,     99.8,
  "30 < Ccr <= 60",     1, "ExI",   1000,  100.0,   100.0,     98.4,
  "30 < Ccr <= 60",     1, "ExI",    500,  100.0,    99.8,     85.1,
  "30 < Ccr <= 60",     1, "ExI",    250,   99.8,    98.4,     35.1,
  "30 < Ccr <= 60",     0, "ExI",   2000,  100.0,   100.0,    100.0,
  "30 < Ccr <= 60",     0, "ExI",   1000,  100.0,   100.0,     99.9,
  "30 < Ccr <= 60",     0, "ExI",    500,  100.0,   100.0,     92.5,
  "30 < Ccr <= 60",     0, "ExI",    250,  100.0,    99.9,     57.8,
  "60 < Ccr <= 90",     0, "InI",   2000,   99.4,    75.9,     59.4,
  "60 < Ccr <= 90",     0, "InI",   1000,   98.9,    69.3,     44.5,
  "60 < Ccr <= 90",     0, "InI",    500,   98.3,    59.4,     23.6,
  "60 < Ccr <= 90",     0, "InI",    250,   96.8,    44.5,      7.6,
  "60 < Ccr <= 90",     0, "ExI",   2000,  100.0,   100.0,    100.0,
  "60 < Ccr <= 90",     0, "ExI",   1000,  100.0,   100.0,     99.7,
  "60 < Ccr <= 90",     0, "ExI",    500,  100.0,   100.0,     86.4,
  "60 < Ccr <= 90",     0, "ExI",    250,  100.0,    99.7,     33.0
)

pta_cmp <- dplyr::inner_join(pta_sim, table4, by = c("band", "crrt", "method", "dose"))
stopifnot(nrow(pta_cmp) == nrow(table4), nrow(pta_cmp) == nrow(pta_arms))

pta_cmp |>
  dplyr::arrange(band, dplyr::desc(crrt), method, dplyr::desc(dose)) |>
  dplyr::mutate(crrt = ifelse(crrt == 1, "On", "Off")) |>
  dplyr::select(band, method, crrt, dose, pub40, pta40, pub100, pta100, pub100x4, pta100x4) |>
  dplyr::rename(
    "Renal function" = band,
    "Infusion" = method,
    "CRRT" = crrt,
    "Dose (mg)" = dose,
    "40% fT>MIC, paper" = pub40,
    "40% fT>MIC, sim" = pta40,
    "100% fT>MIC, paper" = pub100,
    "100% fT>MIC, sim" = pta100,
    "100% fT>4xMIC, paper" = pub100x4,
    "100% fT>4xMIC, sim" = pta100x4
  ) |>
  knitr::kable(digits = 1)
```

For continuous infusion the steady-state concentration is
`Css = (daily dose / 24) / CL`, so a subject attains 100% fT > MIC exactly when
`0.919 x Css > MIC`. That needs only each subject's clearance.

```{r pta-ci}
cl_pta <- sim_pta |>
  dplyr::filter(time == 0) |>
  dplyr::distinct(id, arm_id, cl) |>
  dplyr::left_join(pta_arms, by = "arm_id") |>
  dplyr::filter(method == "InI") |>
  dplyr::mutate(daily_dose = 3 * dose)

pta_ci <- cl_pta |>
  dplyr::mutate(css_free = FU * daily_dose / 24 / cl) |>
  dplyr::group_by(band, crrt, daily_dose) |>
  dplyr::summarise(
    pta100 = 100 * mean(css_free > MIC),
    pta100x4 = 100 * mean(css_free > 4 * MIC),
    .groups = "drop"
  )

pta_ci |>
  dplyr::mutate(crrt = ifelse(crrt == 1, "On", "Off")) |>
  dplyr::arrange(band, dplyr::desc(crrt), dplyr::desc(daily_dose)) |>
  dplyr::rename(
    "Renal function" = band,
    "CRRT" = crrt,
    "Daily dose (mg)" = daily_dose,
    "100% fT>MIC, sim" = pta100,
    "100% fT>4xMIC, sim" = pta100x4
  ) |>
  knitr::kable(digits = 1)
```

The paper's qualitative conclusions are reproduced: every InI and ExI regimen
reaches over 90% PTA for 40% fT > MIC, and every continuous-infusion dose
reaches over 90% PTA for 100% fT > MIC in every renal-function and CRRT
condition. The stricter 100% targets do not match cell by cell. Table 4
changes little with dose (for example 44.5% at 250 mg and 75.9% at 2000 mg for
100% fT > MIC with InI at 60 < Ccr <= 90), whereas the simulation moves from
near 0% to 100% across the same doses: it is higher than Table 4 at the upper
doses and, for 100% fT > 4 x MIC, lower at the lowest ones. An eight-fold dose
increase moving PTA as little as Table 4 shows requires far more
between-subject spread than the printed 7-24% IIV provides, which suggests
residual error was added to the simulated concentrations. With the printed
clearances also giving about twice the exposure of the paper's own fit (see
the VPC above), the stricter-target cells are not expected to match.

```{r pta-gate}
stopifnot(
  # Table 4 / Results: 'For the target of 40% fT > MIC, all three infusion
  # methods of all dosages achieved over 90% PTA regardless of renal function
  # and CRRT.' The simulated cells sit at or near 100%, far from the bound.
  all(pta_cmp$pta40 > 90),
  # Results: 'When administered by CI, all dosages (750-6000 mg) achieved
  # over 90% PTA for 100% fT > MIC'.
  nrow(pta_ci) == 20L,
  all(pta_ci$pta100 > 90)
)
```

# Assumptions and deviations

* **The printed parameters do not reproduce the paper's own fit.** Transcribed
  as printed, the model predicts about twice the concentrations of the paper's
  Figure 5 observed medians and Figure 4 population predictions after the 500
  mg dose that 20 of the 21 patients received. Mass balance on the Figure 5
  median profile (AUC0-inf of roughly 65-70 mg.h/L) implies a pooled clearance
  near 7.4 L/h rather than the 3.65 and 3.81 L/h the paper reports. An earlier
  paper from the same unit (Tanaka et al., Biol Pharm Bull 2017;40:1226-1231,
  doi:10.1248/bpb.b17-00008) studied 12 non-CRRT ICU patients whose
  creatinine clearance summary (mean 76.0 mL/min, range 20.7-155.6) matches the
  non-CRRT stratum here, and reported Bayesian individual clearances of 9.0 +/-
  3.1 L/h, again far above 3.65 L/h. A dose recorded at half its true amount in
  the estimation dataset would halve every clearance and volume and would
  explain the figures, but the paper gives no basis for that correction, so the
  parameters are left exactly as printed. For clinical or exposure work, treat
  the absolute clearance and volume values with caution; the covariate
  structure (stratum-specific creatinine-clearance effects and the additive
  `QE x 0.919` CRRT arm) is unaffected by this issue.
* **IIV order for the two body clearances.** The Results sentence lists the
  inter-individual variability of `CLbody(CRRT)` and `CLbody(non-CRRT)` as 7.3%
  and 22.2% in that order, while Table 3 assigns 7.3% to the non-CRRT and 22.2%
  to the CRRT stratum. The Discussion ("non-CRRT group: 17.0-7.3%, CRRT group:
  29.4-22.2%") agrees with Table 3, which is used here. The shrinkage values in
  the same Results sentence are therefore probably in the same swapped order.
* **Residual error.** The Methods state an additive error model. Table 3 prints
  the final residual row as an estimate of 0.70 (row unit ug/mL) together with
  a percentage of 36.5%, and the Results call 36.5% the residual variability;
  the base-model row (0.016 with 42.3%) and the bootstrap median (0.02) do not
  resolve which is the fitted scale. Both numbers are encoded, as an additive
  SD of 0.70 mg/L plus a proportional SD of 0.365. The widening of the Figure 5
  prediction interval at the 1 h peak supports a concentration-proportional
  component, and its 5th percentile falling below zero at 6-8 h supports an
  additive one. Residual error does not affect the typical-value, NCA or PTA
  sections above.
* **CRRT inputs.** CRRT status is treated as fixed per subject over the 8 h
  sampling interval. The filtrate flow enters as `RRT_CRRT_EFFLUENT_FLOW` in
  mL/h (the paper's QE in L/h times 1000); the paper's QE range (0.6-1.8 L/h)
  equals the sum of its dialysate and replacement-fluid ranges. The 0.919
  sieving coefficient is the literature unbound fraction and was not
  estimated. Per-patient filtrate flows are not published, so the virtual
  cohort draws them uniformly over the reported range.
* **Creatinine clearance** is raw Cockcroft-Gault in mL/min (not BSA-normalized).
  In CRRT patients serum creatinine is also cleared by the circuit, so the
  value reflects combined kidney-plus-circuit function; the model uses the
  CRRT-stratum exponent and median for these patients, as the paper does.
* **Screened but not retained:** body weight (on CL_body, V1, V2) and serum
  albumin (on V1, V2 and CL_CRRT). Body weight on V1 and albumin on CL_CRRT
  entered the full model and were removed in backward elimination (Table 2).
* **Virtual cohort.** Creatinine clearance is drawn from a log-normal
  distribution parameterised with the Table 1 stratum mean and SD and truncated
  to the observed range; for the PTA section
  it is drawn uniformly within each band (the lowest band from 10 mL/min) with
  a filtrate flow of 1.2 L/h, since the paper does not report its simulation
  settings.
