---
title: "S-warfarin PK/PD and INR (Ohara 2014)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{S-warfarin PK/PD and INR (Ohara 2014)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Ohara_2014_warfarin_s"))$reference`
* Description: `r rxode2::rxode(readModelDb("Ohara_2014_warfarin_s"))$description`
* Article (model source): <https://doi.org/10.1371/journal.pone.0105891>
* Article (application / covariate equations): <https://doi.org/10.1097/FPC.0000000000000545>

This model comes from two papers that must be read together.

**Ohara 2014** (PLoS One 9(8):e105891) developed the model in 99 Taiwanese
Chinese patients starting warfarin. It is a three-stage sequential PK/PD model:
S-warfarin plasma concentration (PK) drives inhibition of normal-prothrombin
synthesis (PD-1), and the resulting depletion of normal prothrombin drives INR
(PD-2). Ohara 2014 reports every structural estimate, every fixed constant,
all inter-individual variability and all residual error.

**Shi 2024** (Pharmacogenet Genomics 34:275-284) re-used the model unchanged to
run clinical-trial simulations in Chinese cohorts, comparing the IWPC, Gage and
Miao pharmacogenetic dosing algorithms. Shi 2024 contributes the *explicit
covariate equations* (Table 3 footnotes b-d), which Ohara 2014 states only in
prose, together with the BSA normalisation constant (1.74 m^2) and the baseline
normal-prothrombin centering value (119 mg/L). Shi 2024 deliberately simulated
without unexplained variability, so all the variability in the packaged model
comes from Ohara 2014.

The model file is attributed to Ohara 2014 because the model is entirely theirs;
Shi 2024 changed nothing about it.

# Population

Ohara 2014 studied 99 Chinese patients in Taiwan enrolled in a prospective
randomized trial of genotype-guided (n = 77) versus standard (n = 22) warfarin
initiation (Ohara 2014 Table 1). Mean age was 64.5 +/- 15.2 years, mean body
weight 68.4 +/- 12.4 kg, mean body surface area 1.74 +/- 0.18 m^2, and 39 of 99
(39.4%) were female. Indications were atrial fibrillation (54.5%), stroke
(29.3%), deep vein thrombosis (25.3%) and pulmonary embolism (8.1%). The mean
starting dose was 4.34 +/- 0.98 mg/day of racemic warfarin and the mean
maintenance dose (determined in 89 of 99 patients) was 2.94 +/- 1.35 mg/day,
against a target INR of 2.0-3.0.

The cohort is strongly East-Asian in its pharmacogenetics: VKORC1 -1639 A/A in
81 of 99 (A allele frequency 0.904), CYP2C9*1/*1 in 88 of 99 (\*3 MAF 0.056),
and CYP4F2*3 MAF 0.278. Baseline normal prothrombin (NPT0) was 118.2 +/- 22.1
mg/L and baseline INR 1.05 +/- 0.10. Thirty-five of 99 patients (35%) reached
an INR of 4 or more during induction, which is the over-anticoagulation
phenotype the paper set out to explain.

Shi 2024 applied the model to simulated mainland-Chinese cohorts built from a
660-patient multicentre trial (Shi 2024 Table 1): age 67.4 +/- 10.1 years,
weight 62.2 +/- 12.2 kg, height 161.9 +/- 8.0 cm, VKORC1 -1639 AA 80.2% /
AG 18.0% / GG 1.7%, and CYP2C9 \*1/\*1 92.0% / \*1/\*3 7.6% / \*3/\*3 0.3%.

The same information is available programmatically via the model's `population`
metadata (`rxode2::rxode(readModelDb("Ohara_2014_warfarin_s"))$population`).

# Source trace

Per-parameter origins are recorded as in-file comments next to each `ini()`
entry in `inst/modeldb/specificDrugs/Ohara_2014_warfarin_s.R`. They are
collected here for review. "O" = Ohara 2014, "S" = Shi 2024.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `d/dt(depot)`, `d/dt(central)` (1-cmt, first-order absorption) | n/a | O Eq 1 (written as the analytic ADVAN2/TRANS2 solution; encoded here as the equivalent ODE pair) |
| `d/dt(npt)` (indirect response, synthesis inhibition) | n/a | O Eq 2 |
| `INR` (nonlinear power model) | n/a | O Eq 3 |
| `lfdepot` (F) | `fixed(log(1))` | O Methods, PK paragraph: "F is the bioavailability fixed at 1.0" |
| `lka` | `fixed(log(2))` /h | O Methods, PK paragraph: "Ka is the absorption rate constant fixed at 2 h-1" |
| `lvc` (Vd) | `fixed(log(13.8))` L | O Methods, PK paragraph: "Vd ... of S-warfarin fixed at 13.8 L" (their ref 26) |
| `lcl` (CL(S)) | `log(0.240)` L/h | O Table 2: 240 mL/h (95% CI 220, 260); S Table 3: 0.24 L/h |
| `e_cyp2c9s3_cl` | 0.543 | O Table 2 (95% CI 0.374, 0.712); S Table 3 |
| `e_bsa_cl` | 2.14 | O Table 2 (95% CI 1.12, 3.16); S Table 3 |
| CL(S) covariate equation | `0.24 * 0.543^CYP2C9*3 * (BSA/1.74)^2.14` | S Table 3 footnote b (verbatim) |
| `lic50` | `log(0.0725)` mg/L | O Table 2 (95% CI 0.0631, 0.0819) |
| `lkout` | `log(0.0136)` /h | O Table 2 (95% CI 0.0121, 0.0151) |
| `limax` | `fixed(log(1))` | O Methods, PD-1 paragraph: "IMax ... assumed to be 1.0" |
| `e_vkorc1_ic50` | 2.07 | O Table 2 (95% CI 1.58, 2.56); S Table 3 |
| `e_cyp4f2_ic50` | 1.30 | O Table 2 (95% CI 1.07, 1.53); S Table 3 |
| IC50 covariate equation | `0.0725 * 2.07^VKORC1*2 * 1.30^CYP4F2*3` | S Table 3 footnote c (verbatim) |
| `linrmax` | `fixed(log(5))` | O Table 2 row "INR Max (fixed)"; O Methods, PD-2 paragraph |
| `llambda` | `log(3.48)` | O Table 2 (95% CI 3.30, 3.66) |
| `e_npt0_lambda` | 0.00588 | O Table 2 (95% CI 0.00304, 0.00872) |
| lambda covariate equation | `3.48 * exp(0.00588 * (NPT0 - 119))` | S Table 3 footnote d (verbatim; 119 mg/L is the median NPT0) |
| `kin = kout * NPT_BASE` | n/a | O Methods, PD-1 paragraph: "Kin is expressed as Kout * NPT0" |
| `etalcl` | 0.15920 | O Table 2: omega CL(S) 39.9% -> 0.399^2 |
| `etalic50` | 0.14822 | O Table 2: omega IC50 38.5% -> 0.385^2 |
| `etalkout` | 0.20794 | O Table 2: omega Kout 45.6% -> 0.456^2 |
| `etallambda` | 0.05808 | O Table 2: omega lambda 24.1% -> 0.241^2 |
| `addSd` (Cp(S)) | 0.0697 mg/L | O Table 2, PK residual error (95% CI 0.0676, 0.0718) |
| `addSd_NPT` | 12.2 mg/L | O Table 2, PD-1 residual error (95% CI -16.6, 41.1) |
| `propSd_INR` | 0.247 | O Table 2, PD-2 residual error 24.7% ("relative error model") |
| `NPT_BASE` default | 118.2 mg/L | O Table 1 (mean, n = 99) |
| `INR_BASE` default | 1.05 | O Table 1 (INR0 mean, n = 99) |

# Simulation set-up

All three endpoints of this model (`Cc`, `NPT`, `INR`) are algebraic
observables rather than ODE states, so observation records are tagged with
`dvid = 1` and a blank `cmt`; every algebraic observable is returned on every
solved row regardless.

```{r helpers}
mod <- readModelDb("Ohara_2014_warfarin_s")

# Ohara 2014 reference constants, used throughout as gate targets.
KA_FIX   <- 2      # /h    (Ohara Methods)
VD_FIX   <- 13.8   # L     (Ohara Methods)
CL_REF   <- 0.240  # L/h   (Ohara Table 2, at BSA 1.74, CYP2C9*1/*1)
BSA_REF  <- 1.74   # m^2   (Shi Table 3 footnote b)
IC50_REF <- 0.0725 # mg/L  (Ohara Table 2)
KOUT_REF <- 0.0136 # /h    (Ohara Table 2)
NPT0_REF <- 118.2  # mg/L  (Ohara Table 1)
INR0_REF <- 1.05   #       (Ohara Table 1)

# DuBois BSA, the standard used with warfarin dosing algorithms.
duboisBsa <- function(wtKg, htCm) 0.007184 * wtKg^0.425 * htCm^0.725

# Ohara 2014 typical CL(S) as a closed form, for cross-checking the model.
claim_cl <- function(bsa, cyp2c9s3Carrier = 0) {
  CL_REF * 0.543^cyp2c9s3Carrier * (bsa / BSA_REF)^2.14
}

# Build an event table. `covariates` is a one-row-per-subject data frame that
# must contain `id` plus every covariate column the model needs. Doses are
# S-warfarin doses (see "Assumptions and deviations").
makeEvents <- function(covariates, sDose, ii = 24, untilH, obsTimes) {
  dose <- covariates |>
    dplyr::select(id) |>
    dplyr::mutate(sDose = sDose) |>
    tidyr::expand_grid(time = seq(0, untilH - ii, by = ii)) |>
    dplyr::transmute(
      id, time, amt = sDose, evid = 1L,
      cmt = "depot", dvid = NA_integer_
    )
  obs <- covariates |>
    dplyr::select(id) |>
    tidyr::expand_grid(time = obsTimes) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = NA_character_, dvid = 1L)
  dplyr::bind_rows(dose, obs) |>
    dplyr::left_join(covariates, by = "id") |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

# Deterministic (typical-value) solve: supply zero eta columns and omega = NA
# rather than zeroRe(), which mutates shared model state.
ZERO_ETAS <- list(etalcl = 0, etalic50 = 0, etalkout = 0, etallambda = 0)

solveTypical <- function(events, keep = character()) {
  for (nm in names(ZERO_ETAS)) events[[nm]] <- ZERO_ETAS[[nm]]
  rxode2::rxSolve(mod, events, omega = NA, keep = keep, returnType = "data.frame")
}
```

# Structural gates

These checks are exact: each has a closed-form answer derived from the paper's
own equations, so a failure is unambiguous. They run as `stopifnot()` so the
vignette cannot render if the packaged model drifts.

## Gate 1 - drug-free steady-state hold

Ohara 2014 sets `Kin = Kout * NPT0`, which means the normal-prothrombin pool
must sit exactly at its measured baseline when no warfarin is present, and INR
must sit exactly at `INR_Base`. This is the single most informative structural
check on the PD-1/PD-2 arms.

```{r gate-steady-state}
covOne <- data.frame(
  id = 1L, BSA = 1.77, CYP2C9_S3_COUNT = 0, VKORC1_1639G_COUNT = 0,
  SNP_CYP4F2_RS2108622_T_COUNT = 0, NPT_BASE = NPT0_REF, INR_BASE = INR0_REF
)

drugFree <- makeEvents(covOne, sDose = 0, untilH = 24, obsTimes = seq(0, 24 * 30, by = 6)) |>
  dplyr::filter(evid == 0L) |>
  solveTypical()

stopifnot(
  max(abs(drugFree$NPT - NPT0_REF)) < 1e-8,
  max(abs(drugFree$INR - INR0_REF)) < 1e-10
)

data.frame(
  Check = c("NPT held at NPT_BASE (mg/L)", "INR held at INR_BASE"),
  Target = c(NPT0_REF, INR0_REF),
  `Max absolute deviation over 30 days` = c(
    max(abs(drugFree$NPT - NPT0_REF)), max(abs(drugFree$INR - INR0_REF))
  ),
  check.names = FALSE
) |>
  knitr::kable(caption = "Gate 1. With no warfarin, the turnover pool and INR are pinned at baseline to machine precision, confirming kin = kout * NPT0 (Ohara 2014 Methods, PD-1 paragraph).")
```

## Gate 2 - covariate multipliers reproduce Ohara Table 2 exactly

Each genotype effect in the source papers is a *carriage* multiplier, so a
homozygous variant subject gets the same factor as a heterozygote.

```{r gate-covariates}
covGrid <- expand.grid(
  CYP2C9_S3_COUNT = 0:1, VKORC1_1639G_COUNT = 0:1,
  SNP_CYP4F2_RS2108622_T_COUNT = 0:1
) |>
  dplyr::mutate(
    id = dplyr::row_number(), BSA = BSA_REF,
    NPT_BASE = NPT0_REF, INR_BASE = INR0_REF
  )

covCheck <- makeEvents(covGrid, sDose = 0, untilH = 24, obsTimes = 0) |>
  dplyr::filter(evid == 0L) |>
  solveTypical() |>
  dplyr::left_join(covGrid |> dplyr::select(id, dplyr::everything()), by = "id",
                   suffix = c("", ".cov"))

expectedCl   <- CL_REF * 0.543^covGrid$CYP2C9_S3_COUNT
expectedIc50 <- IC50_REF * 2.07^covGrid$VKORC1_1639G_COUNT *
  1.30^covGrid$SNP_CYP4F2_RS2108622_T_COUNT

stopifnot(
  max(abs(covCheck$cl   - expectedCl))   < 1e-12,
  max(abs(covCheck$ic50 - expectedIc50)) < 1e-12
)

covCheck |>
  dplyr::transmute(
    `CYP2C9*3 carrier` = CYP2C9_S3_COUNT,
    `VKORC1 -1639G carrier` = VKORC1_1639G_COUNT,
    `CYP4F2*3 carrier` = SNP_CYP4F2_RS2108622_T_COUNT,
    `CL(S) (L/h)` = cl,
    `IC50 (mg/L)` = ic50,
    `lambda` = lambda
  ) |>
  knitr::kable(digits = 5, caption = "Gate 2. Derived CL(S) and IC50 at the reference BSA of 1.74 m^2, matching Ohara 2014 Table 2 multipliers (0.543x for CYP2C9*3, 2.07x for VKORC1 -1639G, 1.30x for CYP4F2*3) to machine precision.")
```

The `lambda` column is constant at
`r sprintf("%.4f", 3.48 * exp(0.00588 * (NPT0_REF - 119)))`
because every row uses the same `NPT_BASE` of `r NPT0_REF` mg/L; Shi 2024
Table 3 footnote d gives
`lambda = 3.48 * exp(0.00588 * (NPT0 - 119))`, so at the cohort mean NPT0 the
exponent sits just below its typical value of 3.48.

## Gate 3 - exact end-to-end steady-state identity

Under a constant-rate infusion the whole three-stage model collapses to closed
form, giving an exact check on every layer at once:

* PK (Eq 1): `Cp(S) = R / CL(S)`
* PD-1 (Eq 2), setting `dNPT/dt = 0` with `IMax = 1`:
  `NPT_ss = NPT0 * (1 - Cp / (IC50 + Cp))`
* PD-2 (Eq 3): `INR_ss = INR_Base + 5 * (1 - NPT_ss / NPT0)^lambda`

```{r gate-identity}
INF_CP  <- 0.25                        # target steady-state Cp(S), mg/L
clOne   <- claim_cl(1.77)
infRate <- INF_CP * clOne              # mg/h
infDur  <- 24 * 80                     # h; >> both the 40 h drug and 51 h NPT half-lives

infEvents <- dplyr::bind_rows(
  data.frame(id = 1L, time = 0, amt = infRate * infDur, rate = infRate,
             evid = 1L, cmt = "central", dvid = NA_integer_),
  data.frame(id = 1L, time = seq(0, 24 * 70, by = 6), amt = NA_real_,
             rate = NA_real_, evid = 0L, cmt = NA_character_, dvid = 1L)
) |>
  dplyr::arrange(time, dplyr::desc(evid)) |>
  dplyr::left_join(covOne, by = "id")

infSs <- solveTypical(infEvents) |> dplyr::slice_tail(n = 1)

expCp   <- infRate / clOne
expNpt  <- NPT0_REF * (1 - expCp / (IC50_REF + expCp))
expLam  <- 3.48 * exp(0.00588 * (NPT0_REF - 119))
expInr  <- INR0_REF + 5 * (1 - expNpt / NPT0_REF)^expLam

stopifnot(
  abs(infSs$Cc  / expCp  - 1) < 1e-8,
  abs(infSs$NPT / expNpt - 1) < 1e-6,
  abs(infSs$INR / expInr - 1) < 1e-6
)

data.frame(
  Layer = c("PK - Cp(S) (mg/L)", "PD-1 - NPT (mg/L)", "PD-2 - INR"),
  `Closed form` = c(expCp, expNpt, expInr),
  Simulated = c(infSs$Cc, infSs$NPT, infSs$INR),
  `Relative error` = sprintf("%.1e", c(abs(infSs$Cc / expCp - 1),
                                       abs(infSs$NPT / expNpt - 1),
                                       abs(infSs$INR / expInr - 1))),
  check.names = FALSE
) |>
  knitr::kable(digits = 6, caption = "Gate 3. Under a constant infusion, all three layers of Ohara 2014 Eqs 1-3 match their closed-form steady states to solver precision.")
```

## Gate 4 - off-drug recovery of the NPT pool

After warfarin is withdrawn, normal prothrombin returns to baseline governed by
`Kout`, i.e. a half-life of `log(2) / 0.0136` =
`r sprintf("%.0f", log(2) / KOUT_REF)` h.

```{r gate-recovery}
recEvents <- makeEvents(covOne, sDose = 1.5, ii = 24, untilH = 24 * 21,
                        obsTimes = seq(0, 24 * 60, by = 1))
rec <- solveTypical(recEvents)

# Fit the log-linear return to baseline late enough that residual S-warfarin
# (half-life ~40 h) has decayed far below the turnover mode (half-life ~51 h).
recTail <- rec |>
  dplyr::filter(time >= 24 * 40, time <= 24 * 58) |>
  dplyr::mutate(gap = NPT0_REF - NPT) |>
  dplyr::filter(gap > 1e-9)

kRecovered <- -stats::coef(stats::lm(log(gap) ~ time, data = recTail))[["time"]]
stopifnot(abs(kRecovered - KOUT_REF) / KOUT_REF < 0.05)

data.frame(
  Quantity = c("Recovery rate constant (1/h)", "Recovery half-life (h)"),
  `Ohara 2014 Table 2` = c(KOUT_REF, log(2) / KOUT_REF),
  `Recovered from simulation` = c(kRecovered, log(2) / kRecovered),
  check.names = FALSE
) |>
  knitr::kable(digits = 5, caption = "Gate 4. The off-drug return of NPT to baseline recovers Kout to within a few percent.")
```

The recovery is not a single exponential: residual S-warfarin decays with the
drug elimination rate (`CL/Vd` =
`r sprintf("%.4f", claim_cl(1.77) / VD_FIX)` /h) while the pool refills at
`Kout` = `r KOUT_REF` /h, so a log-linear fit approaches `Kout` from below as
the drug mode dies out. That is why this gate carries a 5% tolerance whereas
Gate 3, which has an exact closed form, is held to solver precision.

```{r recovery-plot}
rec |>
  dplyr::mutate(day = time / 24) |>
  ggplot(aes(day, NPT)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = NPT0_REF, linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = 20, linetype = "dotted") +
  labs(x = "Time (days)", y = "NPT (mg/L)",
       title = "Normal prothrombin: depletion and off-drug recovery",
       subtitle = "S-warfarin 1.5 mg/d for 20 days (dotted line = last dose), then washout",
       caption = "Dashed line = NPT_BASE (118.2 mg/L, Ohara 2014 Table 1).")
```

# Replicating Ohara 2014 Figure 4

Figure 4 of Ohara 2014 predicts the impact of each covariate "in typical
Chinese patients with a BSA of 1.77 m^2 (165 cm and 70 kg) after administration
of racemic warfarin at 3.0 mg/d", i.e. 1.5 mg/d of S-warfarin.

```{r figure-4-setup}
FIG4_BSA     <- 1.77
FIG4_S_DOSE  <- 1.5     # mg/d S-warfarin = racemic 3.0 mg/d
FIG4_DAYS    <- 14
fig4Obs      <- seq(0, 24 * FIG4_DAYS, by = 1)

runScenarios <- function(scen) {
  ev <- makeEvents(scen, sDose = FIG4_S_DOSE, untilH = 24 * FIG4_DAYS,
                   obsTimes = fig4Obs)
  solveTypical(ev, keep = "scenario") |>
    dplyr::mutate(day = time / 24)
}
```

## Panel A - CYP2C9\*3 and BSA on CL(S), seen in Cp(S)

```{r figure-4a}
scenA <- dplyr::bind_rows(
  data.frame(scenario = "CYP2C9*1/*1, BSA 1.77", CYP2C9_S3_COUNT = 0, BSA = FIG4_BSA),
  data.frame(scenario = "CYP2C9*1/*3, BSA 1.77", CYP2C9_S3_COUNT = 1, BSA = FIG4_BSA),
  data.frame(scenario = "CYP2C9*1/*1, BSA 1.60", CYP2C9_S3_COUNT = 0, BSA = 1.60),
  data.frame(scenario = "CYP2C9*1/*1, BSA 1.90", CYP2C9_S3_COUNT = 0, BSA = 1.90)
) |>
  dplyr::mutate(
    id = dplyr::row_number(), VKORC1_1639G_COUNT = 0,
    SNP_CYP4F2_RS2108622_T_COUNT = 0, NPT_BASE = NPT0_REF, INR_BASE = INR0_REF
  )

runScenarios(scenA) |>
  ggplot(aes(day, Cc, colour = scenario)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time (days)", y = "Cp(S) (mg/L)", colour = NULL,
       title = "Figure 4A - CYP2C9*3 and BSA effects on S-warfarin concentration",
       caption = "Replicates Figure 4A of Ohara 2014 (racemic 3.0 mg/d = S 1.5 mg/d).") +
  theme(legend.position = "bottom")
```

Ohara 2014 report that CYP2C9\*3 has "a stronger impact than that of BSA on
CL(S)", and that CL(S) falls about 13% per 0.1 m^2 decrease in BSA. Both are
reproduced: the \*1/\*3 curve is the highest by a wide margin (CL(S) is 0.543x
the wild-type value), while a 0.17 m^2 BSA change moves the curve much less.

```{r figure-4a-check}
clSpan <- c(
  `CYP2C9*1/*3 vs *1/*1` = claim_cl(FIG4_BSA, 1) / claim_cl(FIG4_BSA, 0),
  `BSA 1.60 vs 1.77` = claim_cl(1.60) / claim_cl(FIG4_BSA),
  `BSA 1.80 vs 1.70 (Ohara text: ~13% per -0.1 m^2)` = claim_cl(1.70) / claim_cl(1.80)
)
knitr::kable(data.frame(Comparison = names(clSpan), `CL(S) ratio` = round(clSpan, 3),
                        check.names = FALSE, row.names = NULL),
             caption = "CL(S) ratios implied by the packaged covariate model. Ohara 2014 Results state a 46% CL(S) reduction for CYP2C9*3 carriers and a 13% reduction going from BSA 1.8 to 1.7 m^2.")
```

## Panels B and C - VKORC1\*2 and CYP4F2\*3 on IC50, seen in NPT

```{r figure-4bc}
scenBC <- dplyr::bind_rows(
  data.frame(scenario = "VKORC1 -1639 AA (reference)", VKORC1_1639G_COUNT = 0,
             SNP_CYP4F2_RS2108622_T_COUNT = 0),
  data.frame(scenario = "VKORC1 -1639 AG/GG carrier", VKORC1_1639G_COUNT = 1,
             SNP_CYP4F2_RS2108622_T_COUNT = 0),
  data.frame(scenario = "CYP4F2*3 carrier", VKORC1_1639G_COUNT = 0,
             SNP_CYP4F2_RS2108622_T_COUNT = 1)
) |>
  dplyr::mutate(
    id = dplyr::row_number(), BSA = FIG4_BSA, CYP2C9_S3_COUNT = 0,
    NPT_BASE = NPT0_REF, INR_BASE = INR0_REF
  )

runScenarios(scenBC) |>
  ggplot(aes(day, NPT, colour = scenario)) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time (days)", y = "NPT (mg/L)", colour = NULL,
       title = "Figure 4B/C - VKORC1*2 and CYP4F2*3 effects on normal prothrombin",
       caption = "Replicates Figures 4B and 4C of Ohara 2014.") +
  theme(legend.position = "bottom")
```

Ohara 2014 state that "mutation of VKORC1\*2 had a greater influence than that
of CYP4F2\*3 on IC50" - reproduced here, since carriage of -1639G raises IC50
2.07-fold versus 1.30-fold for CYP4F2\*3. Higher IC50 means less inhibition of
prothrombin synthesis, so the variant curves sit above the reference.

## Panel D - baseline NPT on the NPT-INR relationship

```{r figure-4d}
nptSeq <- seq(0, NPT0_REF, length.out = 200)
npt0Levels <- c(80, 100, 118.2, 140, 160)

fig4d <- lapply(npt0Levels, function(n0) {
  lam <- 3.48 * exp(0.00588 * (n0 - 119))
  data.frame(
    npt0 = n0, lambda = lam,
    fracRemaining = nptSeq / NPT0_REF,
    INR = INR0_REF + 5 * (1 - nptSeq / NPT0_REF)^lam
  )
}) |>
  dplyr::bind_rows() |>
  dplyr::mutate(npt0 = factor(sprintf("NPT0 = %.1f mg/L (lambda %.2f)", npt0, lambda)))

ggplot(fig4d, aes(100 * fracRemaining, INR, colour = npt0)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = c(2, 3), linetype = "dashed", colour = "grey40") +
  scale_x_reverse() +
  labs(x = "NPT remaining (% of NPT0)", y = "INR", colour = NULL,
       title = "Figure 4D - baseline NPT shifts the NPT-INR curve via lambda",
       caption = "Replicates Figure 4D of Ohara 2014; dashed lines mark the INR 2-3 therapeutic range.") +
  theme(legend.position = "bottom")
```

This reproduces the paper's stated direction: "the lower the NPT0, the smaller
the non-linear index lambda leading to a sharp increase in INR in response to
the reduction of NPT after warfarin dosing." A patient with NPT0 = 80 mg/L
reaches INR 2 after far less prothrombin depletion than one with NPT0 = 160
mg/L.

# Replicating Shi 2024 Figure 3a

Shi 2024 simulated 500 virtual Chinese patients per dosing algorithm and report
(Results, "In order to compare the performance...") that "the mean simulated
S-warfarin concentrations for Gage and IWPC dosing cohorts reached as high as
0.38-0.40 mg/l by about 2 weeks, compared to 0.28-0.29 mg/l for the Miao dosing
cohort."

Shi 2024 does not print the numeric doses their algorithms produced, so the
concentration cannot be predicted from a stated dose. The test run here is the
inverse and is sharper: **use the published concentration window to identify
the daily dose it implies, then check that dose against independently reported
Chinese warfarin maintenance doses.** No model parameter is fitted or tuned -
the PK layer is linear in dose, so the implied dose follows by proportion.

This is also the check that settles a convention neither paper states: whether
the model is dosed with the S-warfarin dose or the racemic dose. The two
readings differ by a factor of two and only one of them lands on a dose any
warfarin algorithm produces.

Shi 2024 deliberately simulated without unexplained variability ("We ignored
unexplained variability as it is not possible to attribute it to or correlate it
with any of the covariates"), so the cohort mean below is taken from a
deterministic solve over the covariate distribution only. Ohara 2014's published
inter-individual variability is added back in the following section.

```{r shi-cohort}
set.seed(20260807)
N_ARM <- 200  # per-arm cap

shiCohort <- tibble::tibble(
  id = seq_len(N_ARM),
  AGE = stats::rnorm(N_ARM, 67.4, 10.1),
  WT  = stats::rnorm(N_ARM, 62.2, 12.2),
  HT  = stats::rnorm(N_ARM, 161.9, 8.0)
) |>
  dplyr::mutate(
    WT = pmax(WT, 35), HT = pmax(HT, 140),
    BSA = duboisBsa(WT, HT),
    # Shi 2024 Table 1 genotype frequencies
    VKORC1_1639G_COUNT = sample(0:2, N_ARM, TRUE, c(0.802, 0.180, 0.017)),
    CYP2C9_S3_COUNT    = sample(0:2, N_ARM, TRUE, c(0.920, 0.076, 0.003)),
    # Shi 2024 Table 3 footnote c: all patients assumed CYP4F2*1/*1
    SNP_CYP4F2_RS2108622_T_COUNT = 0,
    # Not reported by Shi 2024; Ohara 2014 Table 1 distributions used instead.
    NPT_BASE = pmax(stats::rnorm(N_ARM, NPT0_REF, 22.1), 40),
    INR_BASE = pmax(stats::rnorm(N_ARM, INR0_REF, 0.10), 0.8)
  )

sprintf("Simulated cohort: BSA mean %.3f m^2 (Shi Table 1 implies %.3f from 62.2 kg / 161.9 cm)",
        mean(shiCohort$BSA), duboisBsa(62.2, 161.9))
```

```{r shi-simulate}
SHI_DAYS <- 42
REF_S_DOSE <- 1.0                       # arbitrary reference; PK is linear in dose
shiObs <- seq(0, 24 * SHI_DAYS, by = 6)

shiEventsRef <- makeEvents(shiCohort, sDose = REF_S_DOSE,
                           untilH = 24 * SHI_DAYS, obsTimes = shiObs)

# Shi 2024's own method: covariate variation only, no unexplained variability.
shiDet <- solveTypical(shiEventsRef) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(day = time / 24)

# Cohort-mean steady-state Cp(S) per 1 mg/day of S-warfarin.
ccPerMgS <- shiDet |>
  dplyr::filter(day >= 28) |>
  dplyr::summarise(m = mean(Cc)) |>
  dplyr::pull(m)

# Confirm dose linearity before scaling by proportion. The PK layer is linear,
# so this is a property of the solver rather than of the cohort; a 10-subject
# subset is enough and keeps the render cheap.
linSub <- dplyr::slice_head(shiCohort, n = 10)
linMean <- function(sDose) {
  solveTypical(makeEvents(linSub, sDose = sDose, untilH = 24 * SHI_DAYS,
                          obsTimes = shiObs)) |>
    dplyr::filter(!is.na(Cc), time >= 24 * 28) |>
    dplyr::summarise(m = mean(Cc)) |>
    dplyr::pull(m)
}
stopifnot(abs(linMean(2 * REF_S_DOSE) / (2 * linMean(REF_S_DOSE)) - 1) < 1e-6)

sprintf("Cohort-mean steady-state Cp(S) = %.4f mg/L per 1 mg/day S-warfarin", ccPerMgS)
```

```{r shi-gate}
SHI_TARGET <- c(0.28, 0.29)   # Shi 2024 Results / Fig 3a, Miao dosing cohort

impliedSDose   <- SHI_TARGET / ccPerMgS          # mg/day S-warfarin
impliedRacemic <- 2 * impliedSDose               # mg/day racemic, if dose = S-dose

# Chinese warfarin maintenance doses reported independently of Shi's Fig 3a:
#   Ohara 2014 Table 1  : 2.94 +/- 1.35 mg/day racemic (n = 89)
#   Shi 2024 Discussion : "2-3 mg/d in the Chinese population"
stopifnot(
  all(impliedRacemic >= 2.0, impliedRacemic <= 3.0),   # S-dose reading is plausible
  all(impliedSDose   <  2.0)                           # racemic reading is not
)

data.frame(
  `Dosing convention` = c("Dose = S-warfarin (half racemic)", "Dose = full racemic"),
  `Implied racemic maintenance dose (mg/day)` = c(
    sprintf("%.2f - %.2f", impliedRacemic[1], impliedRacemic[2]),
    sprintf("%.2f - %.2f", impliedSDose[1], impliedSDose[2])
  ),
  `Consistent with reported Chinese doses?` = c(
    "yes (Ohara 2014 Table 1: 2.94 +/- 1.35; Shi 2024 Discussion: 2-3 mg/d)",
    "no (far below any reported maintenance dose)"
  ),
  check.names = FALSE
) |>
  knitr::kable(caption = "Shi 2024 Figure 3a gate. The published Miao-cohort plateau of 0.28-0.29 mg/L is inverted through the packaged model to identify the daily dose it implies. Only the S-warfarin reading lands on a dose either paper reports.")
```

The half-racemic reading implies a racemic maintenance dose of
`r sprintf("%.2f-%.2f", impliedRacemic[1], impliedRacemic[2])` mg/day, which sits
squarely inside Ohara 2014's observed 2.94 +/- 1.35 mg/day and inside the
"2-3 mg/d in the Chinese population" that Shi 2024's Discussion gives as the
typical empirical starting dose. The full-racemic reading implies
`r sprintf("%.2f-%.2f", impliedSDose[1], impliedSDose[2])` mg/day, which no
warfarin dosing algorithm produces. **The model must therefore be dosed with the
S-warfarin dose**, matching the convention already used in
`vignettes/articles/Hamberg_2007_warfarin_pkpd_pgx.Rmd`.

The Gage and IWPC cohorts in Shi 2024 Figure 3a peak at 0.38-0.40 mg/L, i.e.
`r sprintf("%.0f-%.0f%%", 100 * (0.38 / 0.29 - 1), 100 * (0.40 / 0.28 - 1))`
above the Miao cohort. Because the PK layer is linear in dose, that is a direct
readout of how much the two Western algorithms over-dose this population - the
paper's central finding, and consistent with the over-prediction rates Shi 2024
cites from Xia et al. (46.0% and 51.9% of patients dosed above 120% of actual).

# Population simulation with Ohara 2014 variability

Shi 2024 switched off unexplained variability; the packaged model carries
Ohara 2014's published inter-individual variability, so it can also be used for
variability-aware simulation. The cohort below is dosed at the identified
maintenance dose.

```{r shi-iiv}
S_MAINT <- mean(impliedSDose)   # mg/day S-warfarin, identified above

shiSim <- rxode2::rxSolve(
  mod,
  makeEvents(shiCohort, sDose = S_MAINT, untilH = 24 * SHI_DAYS, obsTimes = shiObs),
  returnType = "data.frame"
) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(day = time / 24)

shiSummary <- shiSim |>
  dplyr::group_by(day) |>
  dplyr::summarise(
    meanCc = mean(Cc), p05 = quantile(Cc, 0.05), p95 = quantile(Cc, 0.95),
    meanInr = mean(INR), inrP05 = quantile(INR, 0.05), inrP95 = quantile(INR, 0.95),
    .groups = "drop"
  )

shiDetSummary <- shiDet |>
  dplyr::group_by(day) |>
  dplyr::summarise(meanCc = mean(Cc) * S_MAINT, .groups = "drop")

ggplot(shiSummary, aes(day, meanCc)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.2) +
  geom_line(linewidth = 0.8) +
  geom_line(data = shiDetSummary, linetype = "dashed", colour = "firebrick") +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = SHI_TARGET[1], ymax = SHI_TARGET[2],
           fill = "steelblue", alpha = 0.35) +
  labs(x = "Time (days)", y = "Cp(S) (mg/L)",
       title = "Figure 3a - S-warfarin concentration in a Chinese cohort",
       subtitle = "Solid = mean with Ohara IIV, ribbon = 5th-95th percentile, dashed red = Shi's no-IIV mean; blue band = Shi 2024 target",
       caption = "Replicates Figure 3a of Shi 2024 at the identified Miao-cohort maintenance dose.")
```

The dashed red line is Shi 2024's own no-variability construction and sits in
the published band by construction. The solid line runs about
`r sprintf("%.0f%%", 100 * (mean(shiSummary$meanCc[shiSummary$day >= 28]) / mean(shiDetSummary$meanCc[shiDetSummary$day >= 28]) - 1))`
higher because clearance is log-normally distributed and concentration is
proportional to its reciprocal, so the population *mean* concentration exceeds
the typical-subject concentration. This is why Shi 2024's decision to zero the
variability matters when comparing against their figures.

```{r shi-inr}
ggplot(shiSummary, aes(day, meanInr)) +
  geom_ribbon(aes(ymin = inrP05, ymax = inrP95), alpha = 0.2) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = c(2, 3), linetype = "dashed", colour = "grey40") +
  labs(x = "Time (days)", y = "INR",
       title = "Figure 3b - INR time course at a fixed maintenance dose",
       subtitle = "Dashed lines mark the INR 2-3 therapeutic range",
       caption = "Companion to Shi 2024 Figure 3b. NOTE: no dose titration is applied here, unlike Shi 2024.")
```

Two things are worth reading carefully in this panel. First, INR keeps climbing
for roughly two weeks after Cp(S) has plateaued, because the normal-prothrombin
pool turns over slowly (half-life `r sprintf("%.0f", log(2) / KOUT_REF)` h). That
delay is the whole reason Ohara 2014 needed an indirect-response layer, and it
is why Shi 2024 found over-anticoagulation appearing "by about 2 weeks".

Second, this simulation applies a *fixed* dose with no titration, whereas both
source papers adjust doses against measured INR (Shi 2024 Table 2). The
untitrated steady-state INR therefore runs above the therapeutic range even at
the cohort's mean maintenance dose. This is consistent with, not contradicted
by, Ohara 2014 Table 1: their reported mean INR of 2.25 is footnoted "Data are
mean values of all measured INRs", i.e. an average over the whole induction
period that starts near 1.05, and 35% of their patients did reach INR of 4 or
more.

```{r shi-inr-summary}
shiSim |>
  dplyr::group_by(id) |>
  dplyr::summarise(peakInr = max(INR), .groups = "drop") |>
  dplyr::summarise(
    `Median peak INR` = median(peakInr),
    `Fraction reaching INR >= 4` = mean(peakInr >= 4)
  ) |>
  knitr::kable(digits = 3, caption = sprintf("Untitrated simulation at the identified maintenance dose (%.2f mg/d racemic). Ohara 2014 reported 35%% of patients reaching INR >= 4 during induction, under a higher starting dose (4.34 mg/d) and with titration.", 2 * S_MAINT))
```

# PKNCA validation of the S-warfarin arm

The PK layer is a one-compartment model with first-order absorption and no
estimated inter-individual variability in `Ka` or `Vd`, so its NCA parameters
have exact closed forms. Comparing PKNCA output against those closed forms
validates both the packaged model and the NCA set-up itself.

```{r pknca-cohort}
ncaScen <- dplyr::bind_rows(
  data.frame(treatment = "CYP2C9*1/*1, BSA 1.74", CYP2C9_S3_COUNT = 0, BSA = 1.74),
  data.frame(treatment = "CYP2C9*1/*3, BSA 1.74", CYP2C9_S3_COUNT = 1, BSA = 1.74),
  data.frame(treatment = "CYP2C9*1/*1, BSA 1.60", CYP2C9_S3_COUNT = 0, BSA = 1.60),
  data.frame(treatment = "CYP2C9*1/*1, BSA 1.90", CYP2C9_S3_COUNT = 0, BSA = 1.90)
) |>
  dplyr::mutate(
    id = dplyr::row_number(), VKORC1_1639G_COUNT = 0,
    SNP_CYP4F2_RS2108622_T_COUNT = 0, NPT_BASE = NPT0_REF, INR_BASE = INR0_REF
  )

S_SINGLE_DOSE <- 1.47  # mg S-warfarin = racemic 2.94 mg (Ohara 2014 Table 1)
ncaObs <- sort(unique(c(seq(0, 24, by = 0.25), seq(24, 24 * 14, by = 2))))

ncaEvents <- makeEvents(ncaScen, sDose = S_SINGLE_DOSE, ii = 24 * 999,
                        untilH = 24 * 999, obsTimes = ncaObs)
ncaSim <- solveTypical(ncaEvents, keep = "treatment")
```

```{r pknca}
simNca <- ncaSim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time-zero record per subject (extravascular pre-dose Cc = 0).
simNca <- dplyr::bind_rows(
  simNca,
  simNca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

concObj <- PKNCA::PKNCAconc(simNca, Cc ~ time | treatment + id)

doseDf <- ncaEvents |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, treatment)
doseObj <- PKNCA::PKNCAdose(doseDf, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

ncaRes <- PKNCA::pk.nca(PKNCA::PKNCAdata(concObj, doseObj, intervals = intervals))
```

```{r pknca-compare}
# Closed-form NCA for a 1-compartment, first-order-absorption model with F = 1.
closedForm <- ncaScen |>
  dplyr::mutate(
    cl   = claim_cl(BSA, CYP2C9_S3_COUNT),
    ke   = cl / VD_FIX,
    tmax = log(KA_FIX / ke) / (KA_FIX - ke),
    cmax = (S_SINGLE_DOSE * KA_FIX / (VD_FIX * (KA_FIX - ke))) *
      (exp(-ke * tmax) - exp(-KA_FIX * tmax)),
    aucinf.obs = S_SINGLE_DOSE / cl,
    half.life  = log(2) / ke
  ) |>
  dplyr::select(treatment, cmax, tmax, aucinf.obs, half.life)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = ncaRes,
  reference = closedForm,
  by        = "treatment",
  units     = c(cmax = "mg/L", tmax = "h", aucinf.obs = "mg*h/L", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated NCA versus the exact closed-form solution of Ohara 2014 Eq 1. * marks a >20% difference. The reference column here is an analytic identity, not a published table - neither source paper reports NCA parameters.",
  align = c("l", "l", "r", "r", "r")
)
```

Cmax, AUC and half-life agree with the analytic solution to the displayed
precision, which confirms three things at once: the ODE encoding of Ohara 2014
Eq 1 is faithful, the covariate model reaches the integrator (clearance, and
hence AUC and half-life, differ correctly across the four scenarios), and the
PKNCA set-up measures what it claims to. The CYP2C9\*3 scenario has a half-life
`r sprintf("%.2f", 1 / 0.543)` times that of the wild-type scenario, exactly as
a clearance ratio of 0.543 requires at fixed `Vd`.

Tmax is the one row with a visible difference (up to about 5%). That is
observation-grid quantisation, not a model discrepancy: PKNCA reports the
sampled time of the maximum, and the grid here is 0.25 h, so a Tmax of 2.39 h
can only be reported as 2.25 or 2.50 h.

# Assumptions and deviations

- **Dosing is expressed as the S-warfarin dose, i.e. half the racemic dose.**
  Neither Ohara 2014 nor Shi 2024 states this explicitly. It is established
  above by the Shi 2024 Figure 3a gate, and it matches the convention already
  used in `vignettes/articles/Hamberg_2007_warfarin_pkpd_pgx.Rmd`. Users
  supplying racemic doses will over-predict exposure two-fold.
- **`NPT_BASE` and `INR_BASE` are covariates, not parameters.** Ohara 2014
  measured both and fitted no variance term for either, excluding the three
  patients whose NPT0 was missing. Simulations here draw them from the Ohara
  2014 Table 1 distributions (118.2 +/- 22.1 mg/L and 1.05 +/- 0.10); Shi 2024
  reports neither for its Chinese cohorts, so the Ohara distributions were used
  for the Shi replication too. Both draws are truncated below (40 mg/L, 0.8) to
  keep them physiological.
- **Inter-individual variability is read as omega, not as a lognormal CV.**
  Ohara 2014 Table 2 labels the rows "Inter-individual error / omega X (%)", so
  the packaged variances are `(percent / 100)^2`. Reading the same numbers as
  exact lognormal CVs would give variances about 7% smaller (e.g. 0.1478 rather
  than 0.1592 for CL(S)); the difference does not affect any gate above.
- **Genotype effects are carriage flags, not per-allele dosages.** The canonical
  covariate columns are allele counts, and `model()` dichotomises them with
  `> 0` because that is how both source papers encode them (Shi 2024 Table 3
  footnotes b and c). A CYP2C9\*3/\*3 homozygote therefore receives the same
  0.543x clearance multiplier as a \*1/\*3 heterozygote. Ohara 2014 had no
  \*3/\*3 subjects, so the source data cannot distinguish the two encodings.
- **VKORC1 reference orientation differs from the other warfarin models in this
  package.** Ohara 2014 takes -1639 A/A as the reference and reports IC50
  *rising* 2.07-fold in G carriers, whereas `Hamberg_2007_warfarin_s` and
  `Xia_2024_warfarin` take G/G as the reference. The biology is identical; only
  the reported direction differs.
- **`INR` uses a guarded fractional power.** Ohara 2014 Eq 3 raises
  `1 - NPT/NPT0` to a non-integer power. Under this model that base is
  mathematically confined to `[0, 1)`, but a solver step returning a value a few
  ulp above `NPT_BASE` would make the power `NaN`, so `model()` clamps the base
  at zero. This enforces the equation's own domain and changes no result.
- **No dose titration is simulated.** Both source papers adjust doses against
  measured INR (Shi 2024 Table 2 gives the nomogram). The simulations here use
  fixed doses, so the INR panels show the untitrated steady state and run above
  the therapeutic range at maintenance doses. Reproducing Shi 2024's %TTR and
  Kaplan-Meier endpoints would require implementing the IWPC, Gage, Miao and
  Lenzini dose-prediction regressions, which are not reprinted in either source
  paper.
- **Shi 2024's supplementary figure is not used.** Supplementary Figure 1
  (<http://links.lww.com/FPC/B501>) shows observed versus simulated INR profiles
  and carries no parameter values.
- **Ohara 2014 Figure 4 is replicated by construction, not by digitisation.**
  The figure's axis values were not digitised; the panels above reproduce its
  stated scenario (BSA 1.77 m^2, racemic 3.0 mg/d) and the directional claims
  quoted from the Results text, and the underlying multipliers are checked
  exactly against Table 2 in Gate 2.
