---
title: "Apremilast popPK and PASI / sPGA exposure-response in Japanese and non-Japanese psoriasis (Okubo 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Apremilast popPK and PASI / sPGA exposure-response in Japanese and non-Japanese psoriasis (Okubo 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Okubo 2021 reports a population PK (PPK) model of oral apremilast, a
phosphodiesterase 4 inhibitor, fitted to pooled Japanese and non-Japanese
data, followed by three exposure-response (E-R) models that link the
individual steady-state AUC from that PPK model to clinical response. Each is
packaged as its own model:

* `Okubo_2021_apremilast`: the one-compartment PPK model (Table 4).
* `Okubo_2021_apremilast_pasi75`: longitudinal logistic E-R model for PASI-75
  (Table S2).
* `Okubo_2021_apremilast_pasi50`: longitudinal logistic E-R model for PASI-50
  (Table S2).
* `Okubo_2021_apremilast_spga`: longitudinal logistic E-R model for sPGA
  response (score 0 or 1) (Table S2).

```{r models}
nms <- c(
  "Okubo_2021_apremilast", "Okubo_2021_apremilast_pasi75",
  "Okubo_2021_apremilast_pasi50", "Okubo_2021_apremilast_spga"
)
uis <- setNames(lapply(nms, function(n) rxode2::rxode(readModelDb(n))), nms)
```

* Citation: `r uis[[1]]$reference`
* Article: <https://doi.org/10.1111/1346-8138.16068>
  (open access; the supporting information holds Tables S1-S3, Appendices
  S1-S5 and Figures S1-S3)

# Population

The PPK analysis pooled 5704 quantifiable apremilast concentrations from 517
subjects in nine studies (Okubo 2021 Results 3.2 and Table 1): 104 Japanese
adults with moderate to severe plaque psoriasis in the phase 2b study
PSOR-011 (apremilast 20 or 30 mg twice daily), 233 non-Japanese psoriasis
patients in the phase 2b PSOR-005 (10, 20 or 30 mg twice daily) and phase 3
PSOR-008/ESTEEM 1 (30 mg twice daily) studies, and 180 non-Japanese subjects
in six phase 1 studies (mostly healthy adults; one study in patients with
psoriasis, psoriatic arthritis or rheumatoid arthritis on methotrexate).
Study-level baseline demographics are in Okubo 2021 Table 3: Japanese
patients averaged 51.4 years and 70.2 kg (75% male), while the non-Japanese
psoriasis patients averaged 46.5-49.1 years and about 95 kg (63-69% male).
Overall 359 of 517 subjects (69%) were male.

The E-R models used 9087 PASI-75 / PASI-50 observations from 1433 subjects
and 9094 sPGA observations from 1442 subjects in PSOR-005, PSOR-008/ESTEEM 1
and PSOR-011, over weeks 2-24, 2-16 and 2-40 respectively (Methods 2.5,
Results 3.3).

```{r population}
str(uis[["Okubo_2021_apremilast"]]$population)
```

# Source trace

Every `ini()` value carries an in-file comment naming its source; the table
collects them.

| Model | Parameter / equation | Value | Source location |
|-------|----------------------|-------|-----------------|
| PPK | `lcl` (CL/F, healthy female non-Japanese, 45 y) | log(9.25) L/h | Table 4 |
| PPK | `lvc` (Vc/F at 82 kg) | log(115) L | Table 4 |
| PPK | `lka` (Ka) | log(1.83) 1/h | Table 4 |
| PPK | `ltlag` (lag time) | log(0.290) h | Table 4 |
| PPK | `e_dis_psoriasis_cl` | 0.834 | Table 4, footnote a |
| PPK | `e_sexm_cl` (male) | 1.25 | Table 4, footnote b |
| PPK | `e_race_japanese_cl` | 1.17 | Table 4, footnote c |
| PPK | `e_age_cl` on (AGE/45) | -0.148 | Table 4, footnote d |
| PPK | `e_wt_vc` on (WT/82) | 0.591 | Table 4, footnote e |
| PPK | `etalcl`, `etalvc`, `etalka` | 38.0%, 27.1%, 83.4% CV | Table 4 (converted with log(1 + CV^2)) |
| PPK | `propSd`, `addSd` | 0.365, 0.658 ng/mL | Table 4 'Error model' |
| PPK | One compartment, first-order absorption with lag | n/a | Results 3.2; Appendix S2-S3 |
| PASI-75 | `bsl_pbo`, `asym_pbo`, `lkpbo` | -7.83, 5.38, log(0.183) 1/week | Table S2 |
| PASI-75 | `emax`, `lauc50` | 3.38, `log(1733)` `ng*h/mL` | Table S2 |
| PASI-50 | `bsl_pbo`, `asym_pbo`, `lkpbo` | -5.84, 4.50, log(0.282) 1/week | Table S2 |
| PASI-50 | `emax`, `lauc50` | 2.91, `log(1656)` `ng*h/mL` | Table S2 |
| sPGA | `bsl_pbo`, `asym_pbo`, `lkpbo` | -8.81, 5.87, log(0.579) 1/week | Table S2 |
| sPGA | `emax`, `lauc50` | 4.58, `log(1110)` `ng*h/mL` | Table S2 |
| sPGA | `e_race_japanese_bsl_pbo`, `_asym_pbo`, `_emax` | 0.682, 0.716, 0.661 | Table S2 'If Japanese' rows |
| sPGA | `lkdrug` (drug-effect onset) | log(0.0540) 1/week | Not printed; back-solved from Table S1 and Results 3.3 (see below) |
| E-R | logit = intercept + placebo x (1 - exp(-K t)) + Emax x AUC / (E50 + AUC) | n/a | Appendix S5; Table S2 definitions |

# Population PK

## Covariate effects quoted in the text

Results 3.2 translates each covariate effect into a percentage change. These
follow from the encoded coefficients alone, so they are checked to the
rounding of the printed percentages.

```{r covariate-claims}
th <- uis[["Okubo_2021_apremilast"]]$theta
claims <- tibble::tribble(
  ~claim, ~model, ~paper,
  "Vc/F at 45.5 kg vs 82 kg", (45.5 / 82)^th[["e_wt_vc"]], 1 - 0.29,
  "Vc/F at 102 kg vs 82 kg", (102 / 82)^th[["e_wt_vc"]], 1 + 0.14,
  "CL/F at 19 y vs 45 y", (19 / 45)^th[["e_age_cl"]], 1 + 0.14,
  "CL/F at 80 y vs 45 y", (80 / 45)^th[["e_age_cl"]], 1 - 0.08,
  "CL/F psoriasis vs healthy", th[["e_dis_psoriasis_cl"]], 1 - 0.17,
  "CL/F male vs female", th[["e_sexm_cl"]], 1.25,
  "CL/F Japanese vs non-Japanese", th[["e_race_japanese_cl"]], 1.17
)
knitr::kable(claims, digits = 3, caption = "Covariate ratios: model vs Okubo 2021 Results 3.2.")
# The printed percentages are rounded to whole percent.
stopifnot(all(abs(claims$model - claims$paper) < 0.006))
```

## Typical-value steady state

For a linear one-compartment model the steady-state AUC over one dosing
interval is exactly dose / CL. A typical Japanese male psoriasis patient
(51 years, 70 kg) at 30 mg twice daily is solved to steady state and the
simulated AUCtau is compared with that closed form.

```{r typical-ss}
mod_pk <- readModelDb("Okubo_2021_apremilast")
tau <- 12
n_dose <- 28
t_last <- (n_dose - 1) * tau
obs_grid <- t_last + c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 2.5, 3, 4, 5, 6, 8, 10, 12)

typ <- data.frame(WT = 70, AGE = 51, SEXF = 0, RACE_JAPANESE = 1, DIS_PSORIASIS = 1)
ev_typ <- bind_rows(
  data.frame(id = 1L, time = (seq_len(n_dose) - 1) * tau, amt = 30, evid = 1L, cmt = "depot"),
  data.frame(id = 1L, time = obs_grid, amt = 0, evid = 0L, cmt = "central")
) |>
  arrange(time, desc(evid)) |>
  cross_join(typ)
sim_typ <- rxode2::rxSolve(rxode2::zeroRe(mod_pk), events = ev_typ,
                           rtol = 1e-10, atol = 1e-12, returnType = "data.frame")
cl_typ <- sim_typ$cl[1]
auc_typ <- with(sim_typ, sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2))
auc_exact <- 1000 * 30 / cl_typ
c(CL_L_per_h = cl_typ, AUCtau_trapezoid = auc_typ, AUCtau_dose_over_CL = auc_exact)
# 28 doses at a 7.5 h half-life is steady state to < 1e-10; the residual is
# the linear-trapezoid error on the 15-point grid, measured at about 0.05%.
stopifnot(abs(auc_typ / auc_exact - 1) < 0.005)
```

## Virtual cohort

Observed data are not public. Virtual psoriasis cohorts follow the Table 3
study means and CVs for PSOR-011 (Japanese) and PSOR-005 / PSOR-008
(non-Japanese); age and weight are drawn from normal distributions truncated
to plausible adult ranges. 150 subjects per arm.

```{r cohort}
set.seed(20211116)
n_arm <- 150
arms <- tibble::tribble(
  ~treatment, ~dose_mg, ~japanese, ~age_m, ~age_cv, ~wt_m, ~wt_cv, ~male,
  "Japanese 20 mg BID", 20, 1L, 51.4, 0.241, 70.2, 0.182, 0.750,
  "Japanese 30 mg BID", 30, 1L, 51.4, 0.241, 70.2, 0.182, 0.750,
  "PSOR-005 10 mg BID", 10, 0L, 46.5, 0.286, 96.3, 0.223, 0.627,
  "PSOR-005 20 mg BID", 20, 0L, 46.5, 0.286, 96.3, 0.223, 0.627,
  "PSOR-005 30 mg BID", 30, 0L, 46.5, 0.286, 96.3, 0.223, 0.627,
  "ESTEEM 1 30 mg BID", 30, 0L, 49.1, 0.276, 94.9, 0.212, 0.687
)
rtrunc <- function(n, m, s, lo, hi) {
  x <- rnorm(n, m, s)
  while (any(bad <- x < lo | x > hi)) x[bad] <- rnorm(sum(bad), m, s)
  x
}
cohort <- arms |>
  mutate(arm = row_number()) |>
  rowwise() |>
  reframe(
    treatment = treatment, dose_mg = dose_mg, RACE_JAPANESE = japanese,
    AGE = rtrunc(n_arm, age_m, age_m * age_cv, 20, 85),
    WT = rtrunc(n_arm, wt_m, wt_m * wt_cv, 40, 180),
    SEXF = as.integer(runif(n_arm) > male)
  ) |>
  mutate(id = row_number(), DIS_PSORIASIS = 1L)

events <- bind_rows(
  cohort |> cross_join(data.frame(time = (seq_len(n_dose) - 1) * tau)) |>
    mutate(amt = dose_mg, evid = 1L, cmt = "depot"),
  cohort |> cross_join(data.frame(time = obs_grid)) |>
    mutate(amt = 0, evid = 0L, cmt = "central")
) |>
  arrange(id, time, desc(evid))
stopifnot(n_distinct(events$id) == nrow(arms) * n_arm)
```

```{r simulate}
sim <- rxode2::rxSolve(mod_pk, events = events, keep = c("treatment", "dose_mg"),
                       returnType = "data.frame")
stopifnot(!anyNA(sim$Cc))
```

## Figure 3: CL/F in Japanese vs non-Japanese psoriasis patients

```{r figure-3, fig.width = 6, fig.height = 4}
indiv <- sim |>
  distinct(id, treatment, dose_mg, cl, vc) |>
  mutate(population = ifelse(grepl("^Japanese", treatment), "Japanese", "Non-Japanese"))
ggplot(indiv, aes(population, cl)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.2, size = 0.6) +
  labs(x = NULL, y = "CL/F (L/h)",
       caption = "Replicates Figure 3 of Okubo 2021 (virtual cohort).")
indiv |>
  group_by(population) |>
  summarise(median_CL = median(cl), .groups = "drop") |>
  knitr::kable(digits = 2, caption = "Median simulated CL/F by population.")
```

## Figure 4: steady-state profiles

```{r figure-4, fig.width = 7, fig.height = 4}
sim |>
  filter(dose_mg %in% c(20, 30), !grepl("ESTEEM", treatment)) |>
  mutate(
    population = ifelse(grepl("^Japanese", treatment), "Japanese", "Non-Japanese (PSOR-005)"),
    regimen = paste(dose_mg, "mg BID"),
    tad = time - t_last
  ) |>
  group_by(population, regimen, tad) |>
  summarise(
    Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, Q50, colour = population, fill = population)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.15, colour = NA) +
  geom_line() +
  facet_wrap(~regimen) +
  labs(x = "Time after morning dose at steady state (h)",
       y = "Apremilast (ng/mL)", colour = NULL, fill = NULL,
       caption = "Replicates Figure 4 of Okubo 2021: median and 90% interval.")
```

## PKNCA: steady-state exposure

```{r pknca}
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)
dose_df <- events |>
  filter(evid == 1) |>
  select(id, time, amt, treatment)
conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)
intervals <- data.frame(
  start = t_last, end = t_last + tau,
  cmax = TRUE, cmin = TRUE, tmax = TRUE, auclast = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

Okubo 2021 Table 5 gives the model-predicted geometric means of AUCtau,ss,
Cmax,ss and Cmin,ss in the E-R populations; these are the most direct check
of the encoded model.

```{r nca-compare-table5}
table5 <- tibble::tribble(
  ~treatment, ~auclast, ~cmax, ~cmin,
  "Japanese 20 mg BID", 1939.2, 240.3, 87.2,
  "Japanese 30 mg BID", 2727.2, 341.8, 120.3,
  "PSOR-005 10 mg BID", 1098.9, 124.0, 58.7,
  "PSOR-005 20 mg BID", 2185.2, 247.7, 115.6,
  "PSOR-005 30 mg BID", 3337.2, 376.7, 178.3,
  "ESTEEM 1 30 mg BID", 3352.4, 377.0, 179.9
)
cmp5 <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = table5,
  by = "treatment",
  params = c("auclast", "cmax", "cmin"),
  units = c(auclast = "ng*h/mL", cmax = "ng/mL", cmin = "ng/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp5, caption = "Simulated (median) vs Okubo 2021 Table 5 (model-predicted geometric mean). * differs by >20%.")
```

```{r nca-gate}
nca_df <- as.data.frame(nca_res)
auc_med <- nca_df |>
  filter(PPTESTCD == "auclast") |>
  group_by(treatment) |>
  summarise(sim = median(PPORRES), .groups = "drop") |>
  left_join(table5 |> select(treatment, ref = auclast), by = "treatment") |>
  mutate(pct_diff = 100 * (sim / ref - 1))
auc_med
# Structural: a wrong CL, covariate factor or unit moves AUC by tens of
# percent. The cohort is a demographic approximation, so the arm medians
# are compared to the centre of Table 5, not to its extremes.
stopifnot(abs(median(auc_med$pct_diff)) < 10, max(abs(auc_med$pct_diff)) < 20)
```

Okubo 2021 Table 2 also reports the observed non-compartmental week-20
values in the intensive-sampling Japanese subset (n = 21 and 20). They are
shown for context only: the subset is small, AUCtau rises less than
dose-proportionally between 20 and 30 mg in those data, and the trapezoid
AUCtau there extrapolates the 12 h concentration from 8 h. The model's own
predictions in Table 5 sit below the observed 20 mg Cmax as well (240 vs 304
ng/mL), so the starred 20 mg Cmax row reflects that subset rather than the
encoding.

```{r nca-compare-table2}
table2 <- tibble::tribble(
  ~treatment, ~auclast, ~cmax, ~cmin, ~tmax,
  "Japanese 20 mg BID", 1957, 304, 90.8, 2.03,
  "Japanese 30 mg BID", 2397, 374, 104, 2.00
)
cmp2 <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = table2,
  by = "treatment",
  params = c("auclast", "cmax", "cmin", "tmax"),
  units = c(auclast = "ng*h/mL", cmax = "ng/mL", cmin = "ng/mL", tmax = "h"),
  tolerance_pct = 20
)
cmp2 <- cmp2[grepl("^Japanese", cmp2$treatment), ]
knitr::kable(cmp2, caption = "Simulated (median) vs Okubo 2021 Table 2 (observed NCA, Japanese, week 20). * differs by >20%.")
```

# Exposure-response

## Week-16 probabilities (Table S1 and Results 3.3)

The E-R models are logistic in the logit
`intercept + placebo x (1 - exp(-K t)) + Emax x AUC / (E50 + AUC)` with `t`
in weeks and `AUC` the individual AUCtau,ss. Okubo 2021 prints 29 week-16
response probabilities at fixed AUC values (Table S1 and Results 3.3). The
models are deterministic given the AUC, so each printed value is reproduced
to its one-decimal rounding.

```{r er-printed}
printed <- tibble::tribble(
  ~endpoint, ~RACE_JAPANESE, ~AUC_APREMILAST, ~paper_pct, ~source,
  "pasi75", 0L, 0, 6.1, "Table S1 placebo",
  "pasi75", 0L, 1065.4, 19.0, "Results 3.3 / Table S1",
  "pasi75", 0L, 2030.6, 28.6, "Results 3.3",
  "pasi75", 0L, 3169.2, 36.5, "Results 3.3",
  "pasi75", 0L, 2136.5, 29.5, "Table S1",
  "pasi75", 0L, 3213.4, 36.8, "Table S1",
  "pasi75", 1L, 1840.5, 27.0, "Table S1",
  "pasi75", 1L, 2720.3, 33.8, "Table S1",
  "pasi75", 1L, 1065.4, 19.0, "Table S1",
  "pasi50", 0L, 0, 20.0, "Table S1 placebo",
  "pasi50", 0L, 1065.4, 43.8, "Results 3.3 / Table S1",
  "pasi50", 0L, 2030.6, 55.3, "Results 3.3",
  "pasi50", 0L, 3169.2, 62.8, "Results 3.3",
  "pasi50", 0L, 2136.5, 56.2, "Table S1",
  "pasi50", 0L, 3213.4, 63.0, "Table S1",
  "pasi50", 1L, 1840.5, 53.6, "Table S1",
  "pasi50", 1L, 2720.3, 60.3, "Table S1",
  "spga", 0L, 0, 5.0, "Table S1 placebo",
  "spga", 1L, 0, 14.1, "Table S1 placebo",
  "spga", 0L, 1065.4, 16.2, "Results 3.3 / Table S1",
  "spga", 0L, 2030.6, 22.7, "Results 3.3",
  "spga", 0L, 3169.2, 27.3, "Results 3.3",
  "spga", 0L, 2136.5, 23.2, "Table S1",
  "spga", 0L, 3213.4, 27.5, "Table S1",
  "spga", 1L, 1065.4, 27.9, "Table S1",
  "spga", 1L, 1840.5, 32.9, "Table S1",
  "spga", 1L, 2720.3, 36.3, "Table S1",
  "spga", 1L, 2030.6, 33.8, "Results 3.3",
  "spga", 1L, 3169.2, 37.5, "Results 3.3"
)

er_models <- list(
  pasi75 = readModelDb("Okubo_2021_apremilast_pasi75"),
  pasi50 = readModelDb("Okubo_2021_apremilast_pasi50"),
  spga = readModelDb("Okubo_2021_apremilast_spga")
)
er_out <- c(pasi75 = "prob_pasi75", pasi50 = "prob_pasi50", spga = "prob_spga01")

# Solve one E-R model on a data frame of observation rows (one row per id).
# The E-R models have no random effects by design (Table S2), so rxode2's
# notice about a multi-subject solve without omega is expected and muffled.
muffle_no_omega <- function(w) {
  if (grepl("without 'omega'", conditionMessage(w))) invokeRestart("muffleWarning")
}
solve_er <- function(endpoint, df) {
  ev <- df |> mutate(evid = 0L, amt = 0)
  out <- withCallingHandlers(
    rxode2::rxSolve(er_models[[endpoint]], events = ev, returnType = "data.frame"),
    warning = muffle_no_omega
  )
  if (is.null(out$id)) out$id <- 1L
  stopifnot(nrow(out) == nrow(ev))
  out[[er_out[[endpoint]]]][match(ev$id, out$id)]
}

printed <- printed |>
  group_by(endpoint) |>
  group_modify(function(d, k) {
    d$model_pct <- 100 * solve_er(
      k$endpoint,
      d |> mutate(id = row_number(), time = 16) |> select(id, time, AUC_APREMILAST, RACE_JAPANESE)
    )
    d
  }) |>
  ungroup() |>
  mutate(diff_pp = model_pct - paper_pct)
knitr::kable(printed, digits = 2, caption = "Week-16 response probability (%): model vs Okubo 2021.")
# Printed to 0.1 percentage point; the largest difference is about 0.05.
stopifnot(max(abs(printed$diff_pp)) < 0.1)
```

The sPGA rows are what pin the unprinted drug-effect onset rate. With the
drug term undelayed, the ten treated sPGA rows are overpredicted by 14-34
percentage points. All ten instead require the drug term to stand at the same
fraction of its full size at week 16, whatever the AUC or race:

```{r spga-fraction}
logit <- function(p) log(p / (1 - p))
th_s <- rxode2::rxode(er_models$spga)$theta
frac <- printed |>
  filter(endpoint == "spga", AUC_APREMILAST > 0) |>
  mutate(
    bsl = th_s[["bsl_pbo"]] * th_s[["e_race_japanese_bsl_pbo"]]^RACE_JAPANESE,
    pbo = th_s[["asym_pbo"]] * th_s[["e_race_japanese_asym_pbo"]]^RACE_JAPANESE *
      (1 - exp(-exp(th_s[["lkpbo"]]) * 16)),
    drug_full = th_s[["emax"]] * th_s[["e_race_japanese_emax"]]^RACE_JAPANESE *
      AUC_APREMILAST / (exp(th_s[["lauc50"]]) + AUC_APREMILAST),
    fraction = (logit(paper_pct / 100) - bsl - pbo) / drug_full
  ) |>
  select(RACE_JAPANESE, AUC_APREMILAST, paper_pct, fraction)
knitr::kable(frac, digits = 4, caption = "Fraction of the sPGA drug term needed at week 16.")
range(frac$fraction)
c(onset_fraction_wk16 = 1 - exp(-exp(th_s[["lkdrug"]]) * 16))
stopifnot(diff(range(frac$fraction)) < 0.003)
```

## Figure 5: probability of response at week 16 vs exposure

```{r figure-5, fig.width = 8, fig.height = 3.5}
grid_auc <- seq(0, 6000, by = 50)
fig5 <- tidyr::expand_grid(endpoint = names(er_models), RACE_JAPANESE = 0:1,
                           AUC_APREMILAST = grid_auc) |>
  group_by(endpoint) |>
  group_modify(function(d, k) {
    d$prob <- solve_er(k$endpoint, d |> mutate(id = row_number(), time = 16) |>
                         select(id, time, AUC_APREMILAST, RACE_JAPANESE))
    d
  }) |>
  ungroup() |>
  mutate(
    endpoint = factor(endpoint, c("pasi75", "pasi50", "spga"),
                      c("PASI-75", "PASI-50", "sPGA 0/1")),
    population = ifelse(RACE_JAPANESE == 1, "Japanese", "Non-Japanese")
  )
ggplot(fig5, aes(AUC_APREMILAST, 100 * prob, linetype = population)) +
  geom_line() +
  facet_wrap(~endpoint) +
  labs(x = "AUCtau,ss (ng*h/mL)", y = "Probability of response at week 16 (%)",
       linetype = NULL,
       caption = "Replicates the model curves of Figure 5 of Okubo 2021.")
```

## Figure S3: response over time

Typical-value time courses at the median non-Japanese AUC for each dose
(Results 3.3), for comparison with the population predictions in
Okubo 2021 Figure S3.

```{r figure-s3, fig.width = 8, fig.height = 3.5}
dose_auc <- tibble::tribble(
  ~regimen, ~AUC_APREMILAST,
  "Placebo", 0,
  "10 mg BID", 1065.4,
  "20 mg BID", 2030.6,
  "30 mg BID", 3169.2
)
figS3 <- tidyr::expand_grid(endpoint = names(er_models), dose_auc,
                            time = seq(0, 24, by = 0.5)) |>
  mutate(RACE_JAPANESE = 0L) |>
  group_by(endpoint) |>
  group_modify(function(d, k) {
    d$prob <- solve_er(k$endpoint, d |> mutate(id = row_number()) |>
                         select(id, time, AUC_APREMILAST, RACE_JAPANESE))
    d
  }) |>
  ungroup() |>
  mutate(endpoint = factor(endpoint, c("pasi75", "pasi50", "spga"),
                           c("PASI-75", "PASI-50", "sPGA 0/1")))
ggplot(figS3, aes(time, 100 * prob, colour = regimen)) +
  geom_line() +
  facet_wrap(~endpoint) +
  labs(x = "Time (weeks)", y = "Probability of response (%)", colour = NULL,
       caption = "Compare with Figure S3 of Okubo 2021 (population predictions).")
```

The sPGA curves rise steadily to week 24 rather than reaching their plateau
by week 8 as the PASI curves do, which is the shape Figure S3C shows (30 mg
twice daily: about 6% at week 4, 16% at week 8 and 40% at week 24).

## PPK to E-R: arm-level week-16 response

Individual AUCtau,ss values from the virtual PK cohort (dose / CL) drive the
three E-R models. The mean predicted probability per arm is shown next to
the observed week-16 response rates the Discussion quotes for the 30 mg
twice-daily arms; this is context, not a gate, because the observed rates
come from single trials.

```{r er-chain}
cohort_auc <- indiv |>
  mutate(AUC_APREMILAST = 1000 * dose_mg / cl,
         RACE_JAPANESE = as.integer(population == "Japanese"),
         time = 16)
arm_pred <- lapply(names(er_models), function(ep) {
  cohort_auc |>
    mutate(prob = solve_er(ep, cohort_auc |> select(id, time, AUC_APREMILAST, RACE_JAPANESE)),
           endpoint = ep)
}) |>
  bind_rows() |>
  group_by(treatment, endpoint) |>
  summarise(pred_pct = 100 * mean(prob), .groups = "drop") |>
  tidyr::pivot_wider(names_from = endpoint, values_from = pred_pct)
observed <- tibble::tribble(
  ~treatment, ~obs_pasi75, ~obs_spga,
  "Japanese 30 mg BID", 28.2, 29.6,
  "PSOR-005 30 mg BID", 40.9, 33.0,
  "ESTEEM 1 30 mg BID", 33.1, 21.7
)
arm_pred |>
  left_join(observed, by = "treatment") |>
  dplyr::rename(
    "Arm" = treatment,
    "PASI-75 model (%)" = pasi75, "PASI-50 model (%)" = pasi50,
    "sPGA model (%)" = spga, "PASI-75 observed (%)" = obs_pasi75,
    "sPGA observed (%)" = obs_spga
  ) |>
  knitr::kable(digits = 1, caption = "Week-16 response: mean model prediction over the virtual cohort vs observed rates quoted in the Okubo 2021 Discussion.")
```

# Assumptions and deviations

- **sPGA drug-effect onset rate (`lkdrug`) is back-solved, not printed.**
  Appendix S5 states that the sPGA E-R model has "an Emax-type effect model
  with an exponential delay component", but Table S2 lists no rate for it.
  The maintainers back-solved it from the ten treated week-16 sPGA
  probabilities printed in Table S1 and Results 3.3. Every one of them
  requires the drug term to be 0.5784 of its full size at week 16 (rounding
  interval 0.5783-0.5786, across AUC values from 1065 to 3213 `ng*h/mL` and
  both races), which with a first-order onset `1 - exp(-kdrug t)` gives
  `kdrug = -log(1 - 0.5784) / 16 = 0.0540` per week. Figure S3C supports the
  gradual onset; a constant scaling of the drug term by 0.58 would instead
  overpredict the early weeks (about 18% at week 4 for 30 mg) and
  underpredict week 24. The value happens to sit next to the printed
  K placebo of 0.579 per week; the maintainers treat that as a coincidence of
  digits, since a drug onset at 0.579 per week would be complete by week 16
  and cannot reproduce Table S1.
- **PASI drug effect is undelayed.** Appendix S5 says the PASI models have
  "a placebo parameter with an exponential delay term applied to the drug
  effect term". Taken literally (the placebo onset also multiplying the drug
  term), the week-16 PASI-75 probability at 30 mg would be 33.9%, not the
  printed 36.5%. All 17 printed PASI-75 / PASI-50 probabilities are
  reproduced to rounding with the exponential onset on the placebo term only
  and the Emax drug term at full size, and this form also matches the early
  weeks of Figure S3A-B, so it is the form encoded.
- **CL/F-Vc/F covariance not reported.** The paper estimated a block
  variance on CL/F and Vc/F (Results 3.2) but Table 4 prints only the two
  variances, so the block is encoded as diagonal (correlation 0).
- **BSV scale.** Table 4 prints between-subject variability as a percentage;
  it is read as a CV and converted with `omega^2 = log(1 + CV^2)`. Reading it
  as `100 * omega` instead would change the CL/F variance from 0.135 to
  0.144 and the Ka variance from 0.528 to 0.696.
- **Fixed-zero random effects omitted.** The lag time (Table 4) and every
  E-R parameter (Table S2) carried a between-subject variability fixed to 0;
  these are encoded without an eta. The E-R models therefore have no random
  effects, and their probabilities are typical-subject (population)
  predictions given the AUC.
- **Residual error of the E-R models.** The source likelihood is Bernoulli
  with no residual error; each E-R model carries a placeholder additive
  residual (`addSd_prob_* = 0.001`, held constant) so the nlmixr2
  observation machinery accepts the model. It is not from the source.
- **Disease reference category.** The PPK model codes disease as psoriasis,
  healthy or other (psoriatic or rheumatoid arthritis in one phase 1 study),
  and only the psoriasis effect was retained; `DIS_PSORIASIS = 0` therefore
  covers both healthy and 'other' subjects.
- **Sex coding.** The paper codes male = 1 and prints a 1.25-fold CL/F for
  males. It is encoded on the canonical `SEXF` as `e_sexm_cl^(1 - SEXF)`, so
  the 9.25 L/h typical value is the female value, as in the paper.
- **Subject count.** Results 3.2 gives 517 subjects; the Table 4 caption
  prints n = 533. The population metadata uses 517, which is also the sum of
  the Table 3 study sizes.
- **Virtual cohort.** Age and body weight are normal with the Table 3 study
  means and CVs, truncated to 20-85 years and 40-180 kg; sex is Bernoulli
  with the Table 3 male fraction. Race, disease status and dose are set per
  arm. The paper's E-R populations (Table 5: n = 89-613 per arm) are not
  reproduced in size.
- **Errata.** No correction notice for doi:10.1111/1346-8138.16068 was found
  in Europe PMC as of 2026-10-02.
