---
title: "Propylene glycol in adults and term neonates (Olafuyi 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Propylene glycol in adults and term neonates (Olafuyi 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r models}
ui_adult   <- rxode2::rxode(readModelDb("Olafuyi_2025_propyleneGlycol_adult"))
ui_neonate <- rxode2::rxode(readModelDb("Olafuyi_2025_propyleneGlycol_neonate"))
```

* Citation: `r ui_adult$reference`
* Adult model: `r ui_adult$description`
* Term-neonate model: `r ui_neonate$description`
* Article: <https://doi.org/10.1002/jcph.6150>

Propylene glycol (PG) is a pharmaceutical excipient that is generally regarded
as safe, but which accumulates when its alcohol-dehydrogenase (ADH) mediated
metabolic clearance saturates. Olafuyi and colleagues measured the ADH kinetics
of PG in pooled human liver cytosol, built Simcyp (Version 20) full-PBPK models
for adults and term neonates around those kinetics, and used them to ask at what
dose PG clearance saturates and at what dose PG reaches the plasma level
associated with toxicity (PLAT, 580 mg/L).

## What is packaged here, and what is not

The source model is a whole-body PBPK model built inside a commercial platform.
Its per-organ tissue partition coefficients, organ volumes and blood flows are
platform database outputs and are not published. What the paper *does* publish is
a complete, self-contained **compound layer** -- a steady-state volume of
distribution, an absorption rate constant and fraction absorbed, an in-vitro
Michaelis constant, a renal clearance, an ADH ontogeny function, and the
resulting predicted hepatic and renal clearances -- and that layer fully
determines a one-compartment reduction.

The two packaged models are therefore **compartmental reductions**, not PBPK
models: a single central compartment sized by the published Vss, with parallel
saturable (Michaelis-Menten) hepatic metabolism and linear renal excretion. The
sections below show that this reduction reproduces the source paper's published
clearances, half-lives, clearance-saturation behaviour, and dose-versus-toxicity
conclusions.

# Population

Both models describe simulated virtual populations rather than fitted cohorts.
Each simulation scenario used 200 virtual subjects drawn from the Simcyp
built-in healthy adult population or from the neonatal age group of the built-in
virtual pediatric population (Methods, "PG Physiologically Based Pharmacokinetic
(PBPK) Model Development"). The compound layer was parameterised from PG
physicochemical properties (Table 1) together with ADH kinetics measured in this
study in pooled human liver cytosol (Table 2).

Model qualification used digitised concentration-time profiles from two
published adult clinical studies, and, in term neonates, individual plasma
concentration-time data from a study at University Hospitals Leuven, Belgium
(internal study number B-32220084836) previously reported by De Cock and
colleagues. That Leuven cohort is itself packaged in this library as an
empirical population-PK model, `modellib("DeCock_2012_propyleneGlycol")`, which
makes a useful independent comparator for the neonatal predictions here.

```{r population}
str(ui_neonate$population, max.level = 1)
```

# Source trace

Per-parameter provenance is recorded as an in-file comment beside each `ini()`
entry in `inst/modeldb/specificDrugs/Olafuyi_2025_propyleneGlycol_adult.R` and
`..._neonate.R`. Collected here for review:

| Equation / parameter | Adult value | Term-neonate value | Source location |
|---|---|---|---|
| `lka` (absorption rate constant) | 7.56 /h | 7.56 /h | Table 1, `ka`; Simcyp mechanistic prediction (footnote a) |
| `lfdepot` (fraction absorbed) | 0.99 | 0.99 | Table 1, `fa`; Simcyp mechanistic prediction (footnote a) |
| `lvc` (Vss per body weight) | 0.80 L/kg | 0.40 L/kg | Table 1, `Vss (L/kg)`; adult footnote c (literature average), pediatric from reference 38 with an optimized Kp scalar of 0.44 |
| `lkm` (Michaelis constant) | 1909.1 mg/L | 1909.1 mg/L | Table 2, `Km` = 25.1 mM, converted with the Table 1 molecular weight 76.06 g/mol |
| `lvmax` (maximum metabolic rate) | 8781.9 mg/h | 668.19 mg/h | **Derived** -- see note below. Table 3 predicted hepatic CL times `km` (neonate additionally divided by the ontogeny factor at birth, 0.18) |
| `lcl_renal` (renal clearance) | 3.7 L/h | 0.011 L/h | Table 3, predicted renal CL |
| `beta_vmax` (ADH fraction at birth) | n/a | 0.18 | Equation (1), `Fbirth`; also Table 1, `F ADH activity, neonate` = 18% |
| `t50_vmax` (age at half-maximal ADH) | n/a | 0.9 years | Equation (1), `Age50` |
| `lhill` (ontogeny Hill exponent) | n/a | 1.4 | Equation (1), `n` |
| `etalvmax` (IIV on vmax) | 0.184417 | 0.169646 | Table 3, hepatic CL %CV of 45 (adult) and 43 (neonate); `omega^2 = log(1 + CV^2)` |
| `etalcl_renal` (IIV on renal CL) | 0.051540 | 0.989541 | Table 3, renal CL %CV of 23 (adult) and 130 (neonate) |
| `propSd` (residual error) | fixed 0 | fixed 0 | Not reported; the source is a simulation study with no fitted residual-error model |
| ADH ontogeny equation | n/a | see below | Equation (1) |
| Saturable elimination `- vmax * Cc / (km + Cc)` | applies | applies | Methods, "PG Model Development in Adults"; the Vmax/Km pair of Table 2 entered as Simcyp cytosolic ADH kinetics |
| Linear renal elimination `- cl_renal * Cc` | applies | applies | Methods; renal excretion set to 45% of published total systemic clearance in adults |

**Why `vmax` is derived rather than transcribed.** Table 2 reports the in-vitro
maximum rate as 1.57 nmole/min/mg of cytosolic protein. Converting that to a
whole-body mg/h rate requires the liver weight and the cytosolic-protein-per-gram
-liver scalar behind the platform's in-vitro-to-in-vivo extrapolation, together
with the liver tissue scalar of 11 (Table 1). Those first two are Simcyp system
parameters and appear nowhere in the paper, so the scaling cannot be done from
on-disk sources. It is instead recovered from the paper's own printed output: in
the sub-saturating concentration range the Michaelis-Menten term reduces to
`(vmax / km) * Cc`, so `vmax = CL_hepatic * km`. The adult and neonatal `vmax`
values above are therefore fixed by Table 3's predicted hepatic clearances of
4.6 and 0.063 L/h and by the Table 2 Michaelis constant. No value was tuned to
match any validation target.

## ADH ontogeny, Equation (1)

```{r ontogeny, fig.height = 3.6}
adh_ontogeny <- function(age_years, fbirth = 0.18, age50 = 0.9, n = 1.4) {
  fbirth + (1 - fbirth) * age_years^n / (age50^n + age_years^n)
}

stopifnot(isTRUE(all.equal(adh_ontogeny(0), 0.18)))

tibble(age = seq(0, 10, by = 0.02)) |>
  mutate(f = adh_ontogeny(age)) |>
  ggplot(aes(age, f)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0.18, linetype = "dotted") +
  geom_vline(xintercept = 0.9, linetype = "dotted") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Age (years)", y = "Fraction of adult ADH activity",
       title = "Equation (1): ADH ontogeny",
       caption = paste("Fbirth = 0.18 at birth (dotted horizontal, matching Table 1's",
                       "18%); Age50 = 0.9 years (dotted vertical); n = 1.4."))
```

# Virtual cohort and simulation helpers

The paper reports Vss per kilogram but hepatic and renal clearance as absolute
L/h for the simulated population, and does not publish a body weight for either
virtual population. Reference weights of 70 kg (adult) and 3.5 kg (term neonate)
are used below. That choice is corroborated rather than assumed: it is checked
against the paper's own published elimination half-lives in the next section.

```{r helpers}
PLAT   <- 580   # mg/L, plasma level associated with toxicity (Wilson 2005, per Discussion)
WT_AD  <- 70    # kg
WT_NN  <- 3.5   # kg

# Typical-value (no between-subject variability) multiple-dose IV simulation.
# `omega = NA` suppresses the model's random effects without mutating the
# shared model object the way zeroRe() would.
sim_typical <- function(ui, wt, age = NULL, mgkg, tau, hours, by = 0.25) {
  n_extra <- hours / tau - 1
  stopifnot(n_extra >= 0, isTRUE(all.equal(n_extra, round(n_extra))))
  ev <- if (n_extra > 0) {
    rxode2::et(amt = mgkg * wt, ii = tau, addl = round(n_extra), cmt = "central")
  } else {
    rxode2::et(amt = mgkg * wt, cmt = "central")   # single dose
  }
  ev <- rxode2::et(ev, seq(0, hours, by = by), cmt = "central")
  cov <- if (is.null(age)) data.frame(WT = wt) else data.frame(WT = wt, AGE = age)
  out <- rxode2::rxSolve(ui, ev, cov, omega = NA, returnType = "data.frame")
  # rxSolve omits `id` for a single subject (see known-vignette-failure-patterns #8)
  if (is.null(out$id)) out$id <- 1L
  out
}
```

# Validation 1: clearance, half-life, and the metabolic/renal split (Table 3, Figure 2)

A single sub-saturating dose (0.75 mg/kg, the lowest dose the paper simulated)
puts the model in its linear range, where NCA recovers the total systemic
clearance and terminal half-life. These are compared against Table 3 and against
the half-life ranges quoted in the Introduction (2-5 h in adults, 10-31 h in
neonates).

```{r nca-linear}
nca_one <- function(ui, wt, age = NULL, arm, hours, by) {
  s <- sim_typical(ui, wt, age, mgkg = 0.75, tau = hours, hours = hours, by = by)
  conc <- s |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::transmute(id, time, Cc, arm = arm)
  conc <- dplyr::bind_rows(
    conc,
    conc |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
  ) |>
    dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
    dplyr::arrange(id, time)
  dose <- data.frame(id = 1L, time = 0, amt = 0.75 * wt, arm = arm)
  list(conc = conc, dose = dose)
}

parts <- list(
  nca_one(ui_adult,   WT_AD, NULL, "Adults",        hours = 72,  by = 0.05),
  nca_one(ui_neonate, WT_NN, 0,    "Term neonates", hours = 200, by = 0.10)
)
conc_all <- dplyr::bind_rows(lapply(parts, `[[`, "conc"))
dose_all <- dplyr::bind_rows(lapply(parts, `[[`, "dose"))

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_all, Cc ~ time | arm + id),
  PKNCA::PKNCAdose(dose_all, amt ~ time | arm + id),
  intervals = data.frame(start = 0, end = Inf,
                         cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE,
                         half.life = TRUE, cl.obs = TRUE)
))

published_cl <- tibble::tribble(
  ~arm,             ~cl.obs, ~half.life,
  "Adults",           8.5,      3.5,
  "Term neonates",    0.083,   20.5
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published_cl,
  by        = "arm",
  params    = c("cl.obs", "half.life"),
  units     = c(cl.obs = "L/h", half.life = "h"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = paste(
    "Simulated versus published PG clearance and half-life.",
    "Reference clearances are the predicted values of Table 3;",
    "reference half-lives are the midpoints of the ranges quoted in the",
    "Introduction (2-5 h in adults, 10-31 h in neonates), which the paper",
    "reports as ranges rather than point estimates.",
    "* marks rows differing from the reference by more than 20%."
  )
)
```

Both clearance rows agree closely with Table 3 (adults -2.4%, term neonates
-10.8%, both far inside the 2-fold criterion the paper itself used). The two
half-life rows carry a `*`, but that flag is an artifact of the comparison
rather than a model discrepancy: the paper reports half-life only as a **range**
(2-5 h in adults, 10-31 h in neonates), so the reference column has to use a
range midpoint, and a midpoint is not a prediction target. Nothing was tuned in
response. The substantive test is whether each simulated half-life falls inside
the published range, which is asserted directly:

```{r halflife-assert}
hl <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD == "half.life") |>
  dplyr::select(arm, half.life = PPORRES)
stopifnot(nrow(hl) == 2L)
hl_ad <- hl$half.life[hl$arm == "Adults"]
hl_nn <- hl$half.life[hl$arm == "Term neonates"]
stopifnot(length(hl_ad) == 1L, length(hl_nn) == 1L)
stopifnot(hl_ad >= 2,  hl_ad <= 5)    # Introduction: 2-5 h in adults
stopifnot(hl_nn >= 10, hl_nn <= 31)   # Introduction: 10-31 h in neonates
cat(sprintf("Adult t1/2 = %.2f h (published 2-5 h); neonatal t1/2 = %.2f h (published 10-31 h)\n",
            hl_ad, hl_nn))
```

The paper also reports the split of total clearance into metabolism and renal
excretion (Figure 2 and Results): 55% metabolised / 45% excreted in adults, and
85% metabolised / 15% excreted in neonates from the observed data. Because the
two elimination arms are separately parameterised, the model's split can be read
straight off its parameters at sub-saturating concentrations.

```{r fm-fe}
split_tbl <- function(ui, wt, age, arm) {
  p <- ui$iniDf
  g <- function(nm) p$est[match(nm, p$name)]
  matur <- if (is.null(age)) 1 else adh_ontogeny(age)
  clh <- exp(g("lvmax")) * matur / exp(g("lkm"))
  clr <- exp(g("lcl_renal"))
  tibble(arm = arm,
         `Hepatic CL (L/h)` = clh,
         `Renal CL (L/h)`   = clr,
         `Total CL (L/h)`   = clh + clr,
         `Fm (%)`           = 100 * clh / (clh + clr),
         `Fe (%)`           = 100 * clr / (clh + clr))
}
splits <- dplyr::bind_rows(
  split_tbl(ui_adult,   WT_AD, NULL, "Adults"),
  split_tbl(ui_neonate, WT_NN, 0,    "Term neonates")
)
knitr::kable(splits, digits = c(0, 4, 4, 4, 1, 1),
             caption = paste("Replicates Figure 2 of Olafuyi 2025:",
                             "mean contributions of hepatic metabolism and renal",
                             "excretion to PG clearance. Published: 55%/45% in",
                             "adults, 85%/15% (observed) in term neonates."))

stopifnot(abs(splits$`Fm (%)`[splits$arm == "Adults"]        - 55) < 2)
stopifnot(abs(splits$`Fm (%)`[splits$arm == "Term neonates"] - 85) < 2)
```

# Validation 2: dose versus clearance (Figure 3)

The paper determined "simulated observed" clearance as AUC divided by dose from
simulated concentration-time profiles over PG doses of 0.75 to 7500 mg/kg per
event, given 6-, 8-, 12-, and 24-hourly, and identified the dose at which that
clearance departs from linearity.

The same sweep is run below. Each arm is dosed for seven days; the apparent
clearance is the total dose delivered on the final day divided by the AUC over
that final 24-hour window, with the AUC computed by PKNCA.

```{r fig3-sim}
dose_grid <- c(0.75, 25, 100, 200, 400, 800, 1200, 2000, 4000, 7500)
taus      <- c(6, 8, 12, 24)
HOURS     <- 168        # 7 days
WINDOW    <- c(144, 168)

arms <- tidyr::expand_grid(
  pop  = c("Adults", "Term neonates"),
  tau  = taus,
  mgkg = dose_grid
) |>
  dplyr::mutate(arm = sprintf("%s | q%dh | %g mg/kg", pop, tau, mgkg),
                id  = dplyr::row_number())

sweep <- lapply(seq_len(nrow(arms)), function(i) {
  a  <- arms[i, ]
  ui <- if (a$pop == "Adults") ui_adult else ui_neonate
  wt <- if (a$pop == "Adults") WT_AD else WT_NN
  ag <- if (a$pop == "Adults") NULL else 0
  s  <- sim_typical(ui, wt, ag, mgkg = a$mgkg, tau = a$tau, hours = HOURS, by = 0.25)
  s |>
    dplyr::filter(!is.na(Cc), time >= WINDOW[1], time <= WINDOW[2]) |>
    dplyr::transmute(id = a$id, time, Cc, arm = a$arm)
})
sweep <- dplyr::bind_rows(sweep)

# One "dose" row per arm carrying the total dose delivered in the final 24 h.
sweep_dose <- arms |>
  dplyr::mutate(amt = mgkg * dplyr::if_else(pop == "Adults", WT_AD, WT_NN) * (24 / tau),
                time = WINDOW[1]) |>
  dplyr::select(id, time, amt, arm)

sweep_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(sweep, Cc ~ time | arm + id),
  PKNCA::PKNCAdose(sweep_dose, amt ~ time | arm + id),
  intervals = data.frame(start = WINDOW[1], end = WINDOW[2],
                         auclast = TRUE, cmax = TRUE)
))

cl_dose <- as.data.frame(sweep_nca) |>
  dplyr::filter(PPTESTCD %in% c("auclast", "cmax")) |>
  dplyr::select(arm, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::left_join(arms |> dplyr::select(arm, pop, tau, mgkg), by = "arm") |>
  dplyr::mutate(
    daily_dose = mgkg * dplyr::if_else(pop == "Adults", WT_AD, WT_NN) * (24 / tau),
    cl_app     = daily_dose / auclast
  )
stopifnot(nrow(cl_dose) == nrow(arms), !anyNA(cl_dose$cl_app))
```

```{r fig3-plot}
cl_dose |>
  mutate(freq = factor(paste0("q", tau, "h"), levels = paste0("q", taus, "h"))) |>
  ggplot(aes(mgkg, cl_app, colour = freq)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.1) +
  facet_wrap(~pop, scales = "free_y") +
  scale_x_log10() +
  labs(x = "PG dose (mg/kg per event, log scale)",
       y = "Apparent clearance, day-7 dose / AUC (L/h)",
       colour = "Dosing frequency",
       title = "Figure 3 - dose versus clearance relationship",
       caption = paste("Replicates Figure 3 of Olafuyi 2025. Clearance is flat in the",
                       "linear range and falls once ADH-mediated metabolism saturates;",
                       "saturation sets in at lower doses with more frequent dosing, and",
                       "at much lower doses in neonates than in adults."))
```

The paper detected saturation as a 1% departure from the linear-phase regression
across a 200-subject virtual population in GraphPad, which cannot be reproduced
exactly from a typical-value profile. The structural signature can be, and it is
sharper. Because both populations share the same Michaelis constant, a
one-compartment reduction predicts that saturation begins at whatever dose puts
the post-dose concentration at a fixed multiple of `km` -- so the ratio of the
adult to the neonatal threshold should equal the ratio of their volumes per
kilogram. The paper's own once-daily thresholds obey this exactly:

```{r saturation-signature}
km_mgL <- 25.1 * 76.06
sig <- tibble::tribble(
  ~pop,             ~vss_L_per_kg, ~q24_threshold_mgkg,
  "Adults",          0.80,          1200,     # Discussion: saturation above 1200 mg/kg q24h
  "Term neonates",   0.40,           600      # Discussion: saturation above 600 mg/kg q24h
) |>
  mutate(`C0 (mg/L)` = q24_threshold_mgkg / vss_L_per_kg,
         `C0 / Km`   = `C0 (mg/L)` / km_mgL)
knitr::kable(sig, digits = c(0, 2, 0, 0, 3),
             caption = paste("Zero-free-parameter structural check. The paper's published",
                             "once-daily saturation thresholds differ two-fold between",
                             "populations, exactly as their volumes per kilogram do, and",
                             "both correspond to the same initial concentration relative",
                             "to the shared Michaelis constant."))
stopifnot(diff(range(sig$`C0 / Km`)) < 1e-9)
```

The frequency dependence is a second independent check. With a longer half-life,
neonatal exposure accumulates more between doses, so the neonatal thresholds must
fall away faster as dosing frequency increases than the adult ones do. The paper
reports adult thresholds of 1200/1200/1000/800 mg/kg and neonatal thresholds of
600/400/250/200 mg/kg for q24/q12/q8/q6 dosing. Relative to the once-daily value,
the reduction predicts the accumulation-driven fall directly:

```{r saturation-frequency}
acc_ratio <- function(cl, v, tau) 1 / (1 - exp(-cl / v * tau))
pred_rel <- function(cl, v) {
  r <- vapply(taus, function(tt) acc_ratio(cl, v, tt), numeric(1))
  r[taus == 24] / r
}
# Linear-range clearances come from the models themselves (the `splits` table
# above), not from transcribed numbers, so this check stays honest if a
# parameter ever changes.
cl_ad <- splits$`Total CL (L/h)`[splits$arm == "Adults"]
cl_nn <- splits$`Total CL (L/h)`[splits$arm == "Term neonates"]

freq_tbl <- tibble(
  freq = paste0("q", taus, "h"),
  `Published adult (mg/kg)`   = c(800, 1000, 1200, 1200)[match(taus, c(6, 8, 12, 24))],
  `Predicted adult (mg/kg)`   = 1200 * pred_rel(cl_ad, 0.80 * WT_AD),
  `Published neonate (mg/kg)` = c(200, 250, 400, 600)[match(taus, c(6, 8, 12, 24))],
  `Predicted neonate (mg/kg)` = 600 * pred_rel(cl_nn, 0.40 * WT_NN)
)
knitr::kable(freq_tbl, digits = 0,
             caption = paste("Saturation threshold versus dosing frequency, anchored on each",
                             "population's published once-daily threshold and scaled only by",
                             "the reduction's own accumulation ratio. No parameter is fitted."))
```

# Validation 3: plasma levels associated with toxicity (Figures 4 and 5)

The paper's clinical conclusions are stated as discrete, checkable claims about
whether steady-state PG concentrations reach the 580 mg/L PLAT at each of its
simulated doses and frequencies. These are reproduced below.

```{r fig4-sim}
plat_arms <- dplyr::bind_rows(
  tidyr::expand_grid(pop = "Adults",        mgkg = c(0.75, 200, 500, 1000), tau = taus),
  tidyr::expand_grid(pop = "Term neonates", mgkg = c(0.75, 50, 100, 200),   tau = taus)
) |>
  dplyr::mutate(arm = sprintf("%s | %g mg/kg | q%dh", pop, mgkg, tau))

# Neonatal elimination is much slower, so the neonatal arms are run for 10 days
# and the adult arms for 7, with the steady-state peak read off the final day of
# each.
plat_hours <- function(pop) if (pop == "Adults") 168 else 240

plat_sim <- lapply(seq_len(nrow(plat_arms)), function(i) {
  a  <- plat_arms[i, ]
  ui <- if (a$pop == "Adults") ui_adult else ui_neonate
  wt <- if (a$pop == "Adults") WT_AD else WT_NN
  ag <- if (a$pop == "Adults") NULL else 0
  hh <- plat_hours(a$pop)
  sim_typical(ui, wt, ag, mgkg = a$mgkg, tau = a$tau, hours = hh, by = 0.25) |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::transmute(time, t_final = time - (hh - 24), Cc,
                     pop = a$pop, mgkg = a$mgkg, tau = a$tau, arm = a$arm)
}) |>
  dplyr::bind_rows()

plat_sim |>
  filter(t_final >= 0) |>
  mutate(freq = factor(paste0("q", tau, "h"), levels = paste0("q", taus, "h")),
         dose = factor(paste0(mgkg, " mg/kg"),
                       levels = paste0(sort(unique(mgkg)), " mg/kg"))) |>
  ggplot(aes(t_final, Cc, colour = dose)) +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = PLAT, linetype = "dotted", colour = "grey30") +
  facet_grid(freq ~ pop, scales = "free_y") +
  labs(x = "Time within the final 24 h of dosing (h)", y = "PG plasma concentration (mg/L)",
       colour = "Dose per event",
       title = "Figure 4 - predicted PG concentration-time profiles at standard dosing frequencies",
       caption = paste("Replicates Figure 4 of Olafuyi 2025. The dotted line is the PLAT",
                       "of 580 mg/L. Adult panels use 0.75/200/500/1000 mg/kg and",
                       "neonatal panels 0.75/50/100/200 mg/kg, as in the source figure."))
```

```{r fig4-table}
plat_reached <- plat_sim |>
  dplyr::filter(t_final >= 0) |>
  dplyr::group_by(pop, mgkg, tau) |>
  dplyr::summarise(cmax_ss = max(Cc), .groups = "drop") |>
  dplyr::mutate(`Reaches PLAT?` = dplyr::if_else(cmax_ss >= PLAT, "yes", "no"))

plat_reached |>
  dplyr::mutate(Frequency = paste0("q", tau, "h"),
                `Dose (mg/kg per event)` = mgkg,
                `Steady-state Cmax (mg/L)` = round(cmax_ss, 1)) |>
  dplyr::select(Population = pop, `Dose (mg/kg per event)`, Frequency,
                `Steady-state Cmax (mg/L)`, `Reaches PLAT?`) |>
  knitr::kable(caption = paste("Steady-state peak PG concentration versus the 580 mg/L PLAT,",
                               "for every dose and frequency shown in Figure 4."))
```

The paper's own statements about these scenarios (Results, "PG Plasma
Concentration Associated with PG Toxicity") are asserted directly against the
table above, so a regression in the model would fail the render rather than pass
silently.

```{r fig4-assert}
reaches <- function(p, d, tt) {
  v <- plat_reached$`Reaches PLAT?`[plat_reached$pop == p &
                                    plat_reached$mgkg == d &
                                    plat_reached$tau == tt]
  if (length(v) != 1L) stop("no unique row for ", p, " ", d, " mg/kg q", tt, "h")
  v
}
cmax_of <- function(p, d, tt) {
  v <- plat_reached$cmax_ss[plat_reached$pop == p & plat_reached$mgkg == d &
                              plat_reached$tau == tt]
  if (length(v) != 1L) stop("no unique row")
  v
}

# Neonates: "plasma levels of PG were below PLAT if 100 mg/kg PG was administered
# once daily, but PG plasma levels reach PLAT if the same dose was given 12, 8,
# and 6 hourly."
stopifnot(reaches("Term neonates", 100, 24) == "no")
stopifnot(reaches("Term neonates", 100, 12) == "yes")
stopifnot(reaches("Term neonates", 100,  8) == "yes")
stopifnot(reaches("Term neonates", 100,  6) == "yes")

# Neonates: "when 50 mg/kg was administered at the standard dosing regimen, PG
# plasma levels were below PLAT at steady state."
stopifnot(all(vapply(taus, function(tt) reaches("Term neonates", 50, tt), character(1)) == "no"))

# Adults: "200 mg/kg dosing on all standard dosing regimen resulted in plasma
# concentration below the PLAT" ... "while at 500 mg/kg and in all standard
# dosing regimen, PG levels reached PLAT."
stopifnot(all(vapply(taus, function(tt) reaches("Adults", 200, tt), character(1)) == "no"))
stopifnot(all(vapply(taus, function(tt) reaches("Adults", 500, tt), character(1)) == "yes"))

# Adults: at 200 mg/kg "the peak plasma level were close to PLAT on the 6-hourly
# dosing schedules"; neonates: at 50 mg/kg "the 6-hourly dosing schedule resulted
# in peak plasma levels close to PLAT". Both must be below the PLAT but within
# striking distance of it -- the band below is the accuracy actually achieved
# (77% and 93% of the PLAT), not a loose placeholder.
stopifnot(dplyr::between(cmax_of("Adults", 200, 6)       / PLAT, 0.70, 1.00))
stopifnot(dplyr::between(cmax_of("Term neonates", 50, 6) / PLAT, 0.85, 1.00))

# The 0.75 mg/kg reference arm is far below the PLAT in both populations at every
# frequency, as in Figure 4.
stopifnot(all(plat_reached$cmax_ss[plat_reached$mgkg == 0.75] < 0.05 * PLAT))

cat(sprintf(paste0("All %d published PLAT scenarios reproduced.\n",
                   "  Adults 200 mg/kg q6h   = %.0f mg/L (%.0f%% of PLAT, 'close to PLAT')\n",
                   "  Neonates 50 mg/kg q6h  = %.0f mg/L (%.0f%% of PLAT, 'close to PLAT')\n",
                   "  Neonates 100 mg/kg q12h= %.0f mg/L (crosses PLAT), q24h = %.0f mg/L (does not)\n"),
            nrow(plat_reached),
            cmax_of("Adults", 200, 6),       100 * cmax_of("Adults", 200, 6) / PLAT,
            cmax_of("Term neonates", 50, 6), 100 * cmax_of("Term neonates", 50, 6) / PLAT,
            cmax_of("Term neonates", 100, 12), cmax_of("Term neonates", 100, 24)))
```

## Proposed safe daily doses (Figure 5)

The paper concludes with a proposed total daily dose of 100-200 mg/kg/day in
adults and 25-50 mg/kg/day in term neonates, at any standard dosing frequency.
Figure 5 shows those regimens sitting below the PLAT. Here the same regimens are
simulated **with** the between-subject variability that Table 3 reports, using
200 virtual subjects per arm, so the check covers the population spread and not
only the typical subject.

```{r fig5-vpc}
set.seed(20250314)
N_PER_ARM <- 200L

make_arm <- function(pop, daily_mgkg, tau, hours, id_offset) {
  wt <- if (pop == "Adults") WT_AD else WT_NN
  per_dose <- daily_mgkg / (24 / tau) * wt
  subj <- tibble(id = id_offset + seq_len(N_PER_ARM), WT = wt)
  if (pop != "Adults") subj$AGE <- 0
  dosing <- subj |>
    tidyr::crossing(time = seq(0, hours - tau, by = tau)) |>
    mutate(amt = per_dose, evid = 1L, cmt = "central")
  obs <- subj |>
    tidyr::crossing(time = seq(0, hours, by = 1)) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
  bind_rows(dosing, obs) |>
    mutate(pop = pop,
           arm = sprintf("%s | %g mg/kg/day | q%dh", pop, daily_mgkg, tau)) |>
    arrange(id, time, desc(evid))
}

ev_ad <- bind_rows(
  make_arm("Adults", 100, 6, 168, 0L),
  make_arm("Adults", 200, 6, 168, 1000L)
)
ev_nn <- bind_rows(
  make_arm("Term neonates", 25, 6, 240, 0L),
  make_arm("Term neonates", 50, 6, 240, 1000L)
)
stopifnot(!anyDuplicated(unique(ev_ad[, c("id", "time", "evid")])))
stopifnot(!anyDuplicated(unique(ev_nn[, c("id", "time", "evid")])))

vpc_ad <- rxode2::rxSolve(ui_adult,   ev_ad, keep = c("arm", "pop")) |> as.data.frame()
vpc_nn <- rxode2::rxSolve(ui_neonate, ev_nn, keep = c("arm", "pop")) |> as.data.frame()
vpc <- bind_rows(vpc_ad, vpc_nn) |> filter(!is.na(Cc))

vpc |>
  group_by(pop, arm, time) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
            .groups = "drop") |>
  ggplot(aes(time, Q50, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = PLAT, linetype = "dotted", colour = "grey30") +
  facet_wrap(~pop, scales = "free") +
  labs(x = "Time (h)", y = "PG plasma concentration (mg/L)",
       colour = NULL, fill = NULL,
       title = "Figure 5 - proposed total daily doses versus the PLAT",
       caption = paste("Replicates Figure 5 of Olafuyi 2025, using 200 virtual subjects",
                       "per arm and the Table 3 clearance %CVs. Bands are the 5th-95th",
                       "percentiles; the dotted line is the 580 mg/L PLAT. Both proposed",
                       "dose ranges are simulated at the 6-hourly schedule, the worst",
                       "case for peak concentration."))

p95_max <- vpc |>
  group_by(arm, time) |>
  summarise(Q95 = quantile(Cc, 0.95), .groups = "drop") |>
  group_by(arm) |>
  summarise(`95th percentile peak (mg/L)` = round(max(Q95), 1), .groups = "drop") |>
  mutate(`Below PLAT?` = if_else(`95th percentile peak (mg/L)` < PLAT, "yes", "no"))
knitr::kable(p95_max,
             caption = paste("Upper (95th percentile) peak concentration for each proposed",
                             "regimen against the 580 mg/L PLAT."))
stopifnot(all(p95_max$`Below PLAT?` == "yes"))
```

Every proposed regimen stays below the PLAT across the upper 95th percentile of
the simulated population, reproducing the paper's central dosing recommendation.

# Assumptions and deviations

* **The packaged models are compartmental reductions of a platform PBPK model,
  not PBPK models.** The source model's per-organ partition coefficients, organ
  volumes and blood flows are Simcyp database outputs and are not published. The
  reduction keeps everything the paper does publish (Vss, `ka`, `fa`, the
  in-vitro Michaelis constant, renal clearance, the ADH ontogeny function) and
  collapses the whole-body distribution into a single central compartment sized
  by the published Vss. PG is a small, neutral, highly water-soluble molecule
  with a plasma unbound fraction of 0.99 and a Vss close to total body water, so
  distribution is effectively instantaneous and the one-compartment reduction is
  well justified. Its adequacy is demonstrated rather than assumed by the
  half-life, clearance-split, saturation, and PLAT checks above.
* **`vmax` is back-solved, not transcribed.** Table 2's in-vitro
  1.57 nmole/min/mg of cytosolic protein cannot be scaled to a whole-body rate
  without the liver weight and the cytosolic-protein-per-gram-liver scalar, which
  are platform system parameters absent from the paper. `vmax` is instead fixed
  by the identity `vmax = CL_hepatic * km` using Table 3's predicted hepatic
  clearances. No parameter was tuned against a validation target. The paper
  publishes no liver weight or cytosolic-protein scalar for either population, so
  the two `vmax` values are not cross-checked against an independent
  in-vitro-to-in-vivo scaling here; what is checked is that the resulting
  clearances, half-lives and saturation behaviour all reproduce the paper's
  published values.
* **Reference body weights are not published.** 70 kg (adult) and 3.5 kg (term
  neonate) are used. Because Vss is parameterised per kilogram and the paper's
  doses are in mg/kg, the initial concentration after a dose is independent of
  this choice; only the half-life and accumulation depend on it, and both
  reference weights are corroborated by the resulting half-lives falling inside
  the paper's published ranges (2-5 h and 10-31 h respectively).
* **Clearance is not weight-scaled.** The paper reports Vss per kilogram but
  hepatic and renal clearance as absolute L/h for each simulated population, and
  publishes no allometric exponent for either arm. No weight scaling was invented
  for the clearance terms, so the models should be used near their reference
  weights.
* **Equation (2) is documented but not implemented.** The paper derives neonatal
  renal function from `GFR (mL/min) = -19.8 + 89.04*BSA - 7.16*BSA^2`. That
  polynomial parameterises the *population* glomerular filtration rate rather
  than PG renal clearance itself -- PG is extensively reabsorbed, so its renal
  clearance is well below GFR in both populations -- and it is numerically
  unstable at neonatal body surface areas, returning small negative values near
  a term neonate's BSA. The published renal clearances of Table 3 are used
  directly instead.
* **Variability terms are virtual-population output dispersions, not estimated
  random effects.** Table 3's %CVs describe the spread of predicted clearance
  across the 200-subject Simcyp populations, driven by the platform's
  physiological variability. They are encoded as fixed etas on `vmax` and
  `cl_renal` and are labelled as such in both model files. No %CV is reported for
  the volume of distribution, so `vc` carries no eta.
* **No residual error is available.** The source is a simulation study with no
  fitted residual-error model, so `propSd` is fixed at 0. The models are intended
  for simulation, not for estimation against observed data as supplied.
* **The supplement is not on disk.** Section S1 (model development detail),
  Figures S1-S5 and Tables S1-S4 could not be retrieved: the EuropePMC
  supplementary-files endpoint returned HTTP 500, the PMC `/bin/` route now
  serves a CAPTCHA rather than the file, and the Wiley landing page returns HTTP
  403. No parameter used above comes from the supplement -- every value is from
  Table 1, Table 2, Table 3, Equation (1) or the Results text -- but the
  supplement's Table S2 holds additional predicted-versus-observed PK parameters
  that would have supplied further validation targets.
* **Saturation thresholds are compared structurally, not by the paper's
  detection rule.** The paper flagged saturation as a 1% departure from a
  linear-phase regression fitted across a 200-subject virtual population in
  GraphPad. That detection rule cannot be reproduced from a typical-value profile
  and is much more sensitive than the doses it reports, so the comparison above
  instead uses two zero-free-parameter structural consequences of the reduction:
  the constant ratio of threshold concentration to Michaelis constant across
  populations, and the accumulation-driven fall of the threshold with dosing
  frequency.
* **Term neonates only.** Equation (1) is continuous in age, so the neonatal
  model's ADH ontogeny extends across childhood; its distribution and renal
  parameters, however, are term-neonate values. The paper is explicit that
  clinically observed PG concentration-time data are unavailable for older
  pediatric age groups, so the model should be used at neonatal ages.
