---
title: "Nicotine intravenous, oral, buccal and transdermal (Olsson Gisleskog 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nicotine intravenous, oral, buccal and transdermal (Olsson Gisleskog 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(
  theme_gray() +
    theme(plot.caption = element_text(hjust = 0), plot.caption.position = "plot", legend.position = "bottom")
)
```

# Model and source

Olsson Gisleskog et al. (2021) pooled 29 single- and repeated-dose studies of
Nicorette and related nicotine replacement products in 930 healthy smokers and
fitted seven population PK models: one for intravenous nicotine and one each
for oral microtablets, mouth spray, chewing gum, lozenge, inhaler and
transdermal patch. Joint fits across routes did not converge, so the
intravenous model was fitted first and its disposition parameters and their
between-subject variability were then held fixed in the six extravascular
models. The library follows the same structure: seven model files, all
described in this article.

* Citation: `r rxode2::rxode(readModelDb("OlssonGisleskog_2021_nicotine_iv"))$reference`
* Article: <https://doi.org/10.1007/s40262-020-00960-5> (open access)
* Supplementary NONMEM control streams and output for all seven final models
  (Electronic Supplementary Material 7):
  <https://static-content.springer.com/esm/art%3A10.1007%2Fs40262-020-00960-5/MediaObjects/40262_2020_960_MOESM7_ESM.docx>

```{r load-models}
model_names <- c(
  iv = "OlssonGisleskog_2021_nicotine_iv",
  oral = "OlssonGisleskog_2021_nicotine_oral",
  mouthspray = "OlssonGisleskog_2021_nicotine_mouthspray",
  gum = "OlssonGisleskog_2021_nicotine_gum",
  lozenge = "OlssonGisleskog_2021_nicotine_lozenge",
  inhaler = "OlssonGisleskog_2021_nicotine_inhaler",
  transdermal = "OlssonGisleskog_2021_nicotine_transdermal"
)
mods <- lapply(model_names, function(n) rxode2::rxode(readModelDb(n)))

# Every model must be integrated as written: a cl/vc pair must not have been
# turned into an analytic linCmt() solution that would drop the absorption
# chains.
stopifnot(all(vapply(mods, function(m) is.null(m$linCmt), logical(1))))

tibble(
  Model = unname(model_names),
  Description = vapply(mods, function(m) m$description, character(1))
) |>
  knitr::kable()
```

All models share the same disposition: three compartments with clearance and
inter-compartmental flows scaled by `(WT/70)^0.75` and volumes by `(WT/70)^1`.
Concentrations are in ng/mL, doses in mg and time in hours.

Every model also carries the paper's device for residual nicotine from
smoking before the study: a virtual bolus of 1 mg given into `central` at the
start of the washout, whose bioavailability `fcentral` (reported as the
"pre-washout nicotine dose", 3.8-7.3 mg across models) has its own IIV and,
where there were several periods, IOV. In the extravascular models only this
virtual dose enters `central`. In the intravenous model the real infusion
also enters `central`, so the virtual-dose record is flagged with
`VIRTUAL_DOSE = 1`. Omit the virtual dose to simulate a subject with no
residual nicotine; this article does so except where a published figure shows
the washout.

# Population

```{r population}
pop <- lapply(mods, function(m) m$population)
tibble(
  Model = names(pop),
  N = vapply(pop, function(p) p$n_subjects, integer(1)),
  Studies = vapply(pop, function(p) p$n_studies, integer(1)),
  `Age, median (range)` = paste0(vapply(pop, function(p) p$age_median, character(1)), " (", vapply(pop, function(p) p$age_range, character(1)), ")"),
  `Weight, median (range)` = paste0(vapply(pop, function(p) p$weight_median, character(1)), " (", vapply(pop, function(p) p$weight_range, character(1)), ")"),
  `Female (%)` = vapply(pop, function(p) p$sex_female_pct, numeric(1)),
  Doses = vapply(pop, function(p) p$dose_range, character(1))
) |>
  knitr::kable(caption = "Study populations by model (Olsson Gisleskog 2021 Tables 1 and 2).")
```

All subjects were healthy adult smokers (typically 15-20 cigarettes/day),
almost all White, studied at clinical units in Sweden between 1993 and 2012.
The intravenous dataset excluded subjects with hepatic or renal impairment.
Demographics were similar across routes, with slightly older subjects in the
intravenous and oral studies. No covariate other than body weight (as fixed
allometry) was retained in any model.

# Source trace

Every `ini()` value carries an in-file comment naming its source: Tables 3-6
of the paper and the final NONMEM control streams and output files in the
electronic supplementary material (ESM). The table below is generated from
those comments so that it cannot drift from the model files. The disposition
parameters and their IIV are the same fixed values in all six extravascular
models; they are listed once, under the intravenous model, and checked for
identity in code below.

```{r source-trace-helpers}
# Parse the ini() block of a model file into (parameter, value, source) rows.
# Assignment lines are followed by a label() line that carries the source
# comment; eta lines carry it on the same line (or on the closing line of a
# multi-line block).
trace_ini <- function(file) {
  x <- readLines(file)
  start <- grep("^\\s*ini\\(\\{", x)
  end <- grep("^\\s*\\}\\)", x)
  end <- end[end > start][1]
  x <- x[(start + 1):(end - 1)]
  rows <- list()
  cur <- NULL
  eta_lhs <- NULL
  eta_rhs <- ""
  for (ln in x) {
    has_cmt <- grepl("#", ln)
    code <- trimws(sub("#.*$", "", ln))
    cmt <- if (has_cmt) trimws(sub("^[^#]*#", "", ln)) else ""
    if (!is.null(eta_lhs)) {
      eta_rhs <- paste(eta_rhs, code)
      if (has_cmt) {
        rows[[length(rows) + 1]] <- c(eta_lhs, eta_rhs, cmt)
        eta_lhs <- NULL
      }
    } else if (grepl("^[A-Za-z][A-Za-z0-9_.]*\\s*<-", code)) {
      cur <- c(trimws(sub("<-.*$", "", code)), trimws(sub("^[^<]*<-", "", code)))
    } else if (grepl("^label\\(", code) && !is.null(cur)) {
      rows[[length(rows) + 1]] <- c(cur, cmt)
      cur <- NULL
    } else if (grepl("~", code)) {
      lhs <- trimws(sub("~.*$", "", code))
      rhs <- trimws(sub("^[^~]*~", "", code))
      if (has_cmt) {
        rows[[length(rows) + 1]] <- c(lhs, rhs, cmt)
      } else {
        eta_lhs <- lhs
        eta_rhs <- rhs
      }
    }
  }
  out <- as.data.frame(do.call(rbind, rows), stringsAsFactors = FALSE)
  names(out) <- c("parameter", "value", "source")
  out
}

model_file <- function(n) {
  f <- system.file("modeldb", "specificDrugs", paste0(n, ".R"), package = "nlmixr2lib")
  if (!nzchar(f)) stop("model file not found for ", n)
  f
}

trace <- bind_rows(lapply(names(model_names), function(k) {
  trace_ini(model_file(model_names[[k]])) |> mutate(model = k, .before = 1)
}))

# Every ini() entry of every model must have been captured with a source.
n_ini <- vapply(mods, function(m) {
  d <- m$iniDf
  # a block of correlated etas is one source row; count its diagonal entries
  # as one row per block
  n_theta <- sum(!is.na(d$ntheta))
  etas <- d[is.na(d$ntheta), ]
  diag_rows <- etas[etas$neta1 == etas$neta2, ]
  in_block <- diag_rows$neta1 %in% c(etas$neta1[etas$neta1 != etas$neta2], etas$neta2[etas$neta1 != etas$neta2])
  n_theta + sum(!in_block) + as.integer(any(in_block))
}, integer(1))
n_trace <- table(factor(trace$model, levels = names(model_names)))
stopifnot(identical(as.integer(n_trace), unname(n_ini)))
stopifnot(all(nzchar(trace$source)))
```

```{r shared-disposition}
shared <- c("lcl", "lvc", "lq", "lvp", "lq2", "lvp2", "e_wt_cl_q", "e_wt_vc_vp")
ev_models <- setdiff(names(mods), "iv")
iv_ini <- mods$iv$iniDf
for (k in ev_models) {
  d <- mods[[k]]$iniDf
  # Disposition thetas: same values as the IV estimates, and fixed.
  stopifnot(
    isTRUE(all.equal(d$est[match(shared, d$name)], iv_ini$est[match(shared, iv_ini$name)])),
    all(d$fix[match(shared, d$name)])
  )
  # Disposition IIV (CL and the V1/V3/V2 block): same values, fixed.
  eta_names <- c("etalcl", "etalvc", "etalvp2", "etalvp")
  iv_om <- mods$iv$omega[eta_names, eta_names]
  om <- mods[[k]]$omega[eta_names, eta_names]
  stopifnot(isTRUE(all.equal(unname(om), unname(iv_om))))
}
```

```{r source-trace-table}
trace |>
  filter(model == "iv" | !(parameter %in% c(shared, "etalcl", "etalvc + etalvp2 + etalvp"))) |>
  rename(Model = model, Parameter = parameter, `Value (as coded)` = value, Source = source) |>
  knitr::kable(caption = "Source of every ini() value (ESM = electronic supplementary material, NONMEM code and output).")
```

The model equations come from the ESM control streams and the paper:

| Equation | Source |
|----------|--------|
| Three-compartment disposition, allometry `(WT/70)^0.75` on CL/Q2/Q3 and `^1` on V1-V3 | Sect. 2.3.1, Fig. 1; ESM `$PK` (all models) |
| Virtual pre-washout bolus into central, `F = pre-washout dose * exp(eta [+ IOV])` | Sect. 2.3.2; ESM `$PK` `F1`/`F2` on `PREDOSE` records |
| Oral: `depot -> central` (92NNBT005) or `depot -> transit1-3 -> central` (93NNBT007) | Sect. 3.3; ESM oral `$DES` |
| Time-dependent change `P = Pbl (1 + Emax [T^pow/(Start^pow + T^pow) - T^pow/((Start + Duration)^pow + T^pow)])` | Eq. 2; ESM `TFL` (oral, spray, lozenge, inhaler on CL; gum on oral F) |
| Buccal split: `F_buccal (1 - Frsw)` to the buccal cavity, `F_oral Frsw` to the gut, `F_oral = 0.395` | Sect. 2.3.4, Fig. 1; ESM `F1`/`F5` |
| Mouth spray/inhaler: `depot_buccal -> central` (ka); `depot_oral -> transit1 -> transit2 -> central` (ktr) | Sect. 3.4; ESM mouth-spray/inhaler `$DES` |
| Gum/lozenge: `depot -> depot_buccal` (Krel) `-> central` (ka); `depot_oral -> transit1-3 -> central` (Krel, then ktr) | Sect. 3.4; ESM gum/lozenge `$DES` |
| Gum release `Krel = -log(1 - ADOSE/NDOSE)/0.5`, stopping 0.5 h after the dose | Eq. 1; ESM gum `KREL`, `ENDDOSE` |
| Fraction swallowed: logit-normal; spray `Frsw = 0.606 (dose/2)^0.0928`; lozenge logit shifts for Nicorette and 2 mg | Table 5; ESM `TVFRSW`, `EFF_FRSW1/2` |
| Patch: `depot_td -> transit1` (Krel while `t <= Frdur1*16 h`) with `F = FTOT Fr1 / (1 - exp(-Krel Frdur1 16))`; zero-order `FTOT (1 - Fr1)` into `transit1` from 4.06 h to removal; `transit1-3 -> central` | Sect. 2.3.5, 3.5, Fig. 1; ESM transdermal `$PK`/`$DES` (`MTIME`, `D5 = DUR - ALAG5`) |
| Patch CL `x (1 + 0.116)` from 25 h after the first application | Table 6 'CLch24'; ESM `TFLAG` |
| Residual error `sqrt(add^2 + prop^2 IPRED^2)` | Sect. 2.3; ESM `$ERROR` |

# Virtual cohort

The individual data are not public. Simulations use virtual subjects whose
body weight follows the published distribution (median about 72 kg, CV about
16%, range 40-113 kg; Table 2), redrawn rather than clipped when outside the
observed range. At most 200 subjects are simulated per arm.

```{r cohort}
set.seed(2021)
draw_wt <- function(n) {
  wt <- rlnorm(n, log(72), 0.16)
  bad <- wt < 40 | wt > 113
  while (any(bad)) {
    wt[bad] <- rlnorm(sum(bad), log(72), 0.16)
    bad <- wt < 40 | wt > 113
  }
  wt
}
n_arm <- 200L

# One event table for an arm: `doses` holds the dose records (time, cmt, amt,
# rate) given to every subject, `covs` the covariate columns. ids are offset
# so arms can be combined.
make_arm <- function(doses, obs_times, covs, treatment, id_offset) {
  wt <- draw_wt(n_arm)
  bind_rows(lapply(seq_len(n_arm), function(i) {
    obs <- tibble(time = obs_times, evid = 0L, cmt = "central", amt = NA_real_, rate = NA_real_)
    bind_rows(mutate(doses, evid = 1L), obs) |>
      mutate(id = id_offset + i, WT = wt[i])
  })) |>
    bind_cols(as_tibble(covs)[rep(1, n_arm * (nrow(doses) + length(obs_times))), ]) |>
    mutate(treatment = treatment) |>
    arrange(id, time, desc(evid))
}
```

# Intravenous model

## Derived disposition quantities

The paper reports the half-lives and the fraction of the area under the curve
of the three disposition phases (7 min, 57 min and 4.5 h; 0.04, 0.36 and
0.61), a typical clearance of 1.1 L/min and a volume of distribution at steady
state of 4.3 L/kg for a 70-kg subject (Sect. 3.2 and Conclusions). These
follow in closed form from the typical parameters.

```{r iv-derived}
th <- function(m, p) {
  v <- m$iniDf$est[m$iniDf$name == p]
  if (length(v) != 1) stop("no unique parameter ", p)
  v
}
cl <- exp(th(mods$iv, "lcl"))
v1 <- exp(th(mods$iv, "lvc"))
q2 <- exp(th(mods$iv, "lq"))
v2 <- exp(th(mods$iv, "lvp"))
q3 <- exp(th(mods$iv, "lq2"))
v3 <- exp(th(mods$iv, "lvp2"))
K <- matrix(c(
  -(cl + q2 + q3) / v1, q2 / v1, q3 / v1,
  q2 / v2, -q2 / v2, 0,
  q3 / v3, 0, -q3 / v3
), 3, 3)
eg <- eigen(K)
lam <- -eg$values
coef <- eg$vectors[1, ] * solve(eg$vectors, c(1, 0, 0)) / v1
ord <- order(lam, decreasing = TRUE)
lam <- lam[ord]
coef <- coef[ord]
phase <- tibble(
  Phase = c("rapid", "intermediate", "slowest"),
  `Half-life (model)` = c(sprintf("%.1f min", 60 * log(2) / lam[1:2]), sprintf("%.2f h", log(2) / lam[3])),
  `Half-life (paper)` = c("7 min", "57 min", "4.5 h"),
  `AUC fraction (model)` = round((coef / lam) / sum(coef / lam), 3),
  `AUC fraction (paper)` = c(0.04, 0.36, 0.61)
)
knitr::kable(phase, caption = "Disposition phases of the typical 70-kg subject after an IV bolus.")
cat(sprintf("CL = %.2f L/min (paper 1.1); Vss = %.2f L/kg (paper 4.3)\n", cl / 60, (v1 + v2 + v3) / 70))

stopifnot(
  round(60 * log(2) / lam[1]) == 7,
  round(60 * log(2) / lam[2]) == 57,
  round(log(2) / lam[3], 1) == 4.5,
  all(abs((coef / lam) / sum(coef / lam) - c(0.04, 0.36, 0.61)) <= 0.005),
  round(cl / 60, 1) == 1.1,
  round((v1 + v2 + v3) / 70, 1) == 4.3
)
```

## Mass balance

A deterministic typical-value solve must return the whole infused dose:
`CL * AUC(0-inf) = dose`. The same identity is checked for every
extravascular model below, where it also tests the dose split between the
buccal cavity and the gut, the bioavailabilities and the release windows.

```{r mass-balance-helpers}
dense_grid <- sort(unique(c(seq(0, 2, 0.005), seq(2, 12, 0.02), seq(12, 120, 0.1))))
auc_trap <- function(t, c) sum(diff(t) * (head(c, -1) + tail(c, -1)) / 2)
solve_typical <- function(m, doses, covs) {
  obs <- tibble(time = dense_grid, evid = 0L, cmt = "central", amt = NA_real_, rate = NA_real_)
  ev <- bind_rows(mutate(doses, evid = 1L), obs) |>
    mutate(id = 1L) |>
    bind_cols(as_tibble(covs)[rep(1, nrow(doses) + length(dense_grid)), ]) |>
    arrange(time, desc(evid))
  rxode2::rxSolve(rxode2::zeroRe(m), ev,
    returnType = "data.frame",
    rtol = 1e-10, atol = 1e-12, maxsteps = 1e6
  )
}
# Relative error of CL * AUC against the expected systemically available
# amount (mg). Concentrations are ng/mL = ug/L, so AUC/1000 is mg*h/L.
mb_error <- function(sim, expected_mg) auc_trap(sim$time, sim$Cc) / 1000 * cl / expected_mg - 1
mass_balance <- list()
```

```{r iv-mass-balance}
dose_iv <- 0.028 * 70 # 0.028 mg/kg over 10 min (Table 1)
s <- solve_typical(
  mods$iv,
  tibble(time = 0, cmt = "central", amt = dose_iv, rate = dose_iv * 6),
  list(WT = 70, VIRTUAL_DOSE = 0)
)
mass_balance$iv <- mb_error(s, dose_iv)
```

## Replicating Figure 3

Figure 3 of the paper is a VPC of the intravenous data, including the washout
period during which residual nicotine from smoking decays. The simulation
below reproduces the study design: a virtual 1-mg bolus with
`VIRTUAL_DOSE = 1` at the start of the 36-h washout, then 0.028 mg/kg infused
over 10 min.

```{r iv-vpc}
times_iv <- c(seq(-36, -4, by = 4), -1, -0.5, seq(0, 1, by = 1 / 12), 1.5, 2, 3, 4, 6, 8, 10, 12, 16, 24)
ev_iv <- bind_rows(lapply(seq_len(n_arm), function(i) {
  wt <- draw_wt(1)
  d <- 0.028 * wt
  bind_rows(
    tibble(time = -36, evid = 1L, cmt = "central", amt = 1, rate = NA_real_, VIRTUAL_DOSE = 1),
    tibble(time = 0, evid = 1L, cmt = "central", amt = d, rate = d * 6, VIRTUAL_DOSE = 0),
    tibble(time = times_iv, evid = 0L, cmt = "central", amt = NA_real_, rate = NA_real_, VIRTUAL_DOSE = 0)
  ) |>
    mutate(id = i, WT = wt)
})) |>
  arrange(id, time, desc(evid))
sim_iv <- rxode2::rxSolve(mods$iv, ev_iv, returnType = "data.frame")
sim_iv |>
  group_by(time) |>
  summarise(
    p05 = quantile(sim, 0.05), p50 = median(sim), p95 = quantile(sim, 0.95),
    .groups = "drop"
  ) |>
  pivot_longer(-time, names_to = "percentile", values_to = "Cc") |>
  mutate(Cc = pmax(Cc, 0.1)) |>
  ggplot(aes(time, Cc, linetype = percentile)) +
  geom_line() +
  geom_hline(yintercept = 0.2, colour = "grey50") +
  scale_y_log10() +
  scale_linetype_manual(values = c(p05 = "dashed", p50 = "solid", p95 = "dashed")) +
  labs(
    x = "Time after the start of the infusion (h)", y = "Nicotine (ng/mL)",
    title = "Intravenous nicotine: simulated 5th, 50th and 95th percentiles",
    caption = "Replicates Figure 3 of Olsson Gisleskog 2021.\nValues below 0.1 ng/mL shown at 0.1; grey line: LLOQ 0.2 ng/mL."
  )
```

```{r iv-vpc-table}
# Approximate values read from the raster Figure 3 (observed percentiles and
# the proportion of observations below the limit of quantification).
fig3 <- tibble(
  time = c(-24, 1 / 6, 6, 12, 24),
  `Figure 3 median (ng/mL)` = c("1.4", "12", "1.5", "0.6", "< LLOQ"),
  `Figure 3 BQL (%)` = c("0", "0", "0", "30", "70")
)
iv_tab <- sim_iv |>
  mutate(time = round(time, 4)) |>
  filter(time %in% round(fig3$time, 4)) |>
  group_by(time) |>
  summarise(
    `Simulated 5th` = quantile(sim, 0.05), `Simulated median` = median(sim),
    `Simulated 95th` = quantile(sim, 0.95), `Simulated BQL (%)` = 100 * mean(sim < 0.2),
    .groups = "drop"
  ) |>
  inner_join(fig3 |> mutate(time = round(time, 4)), by = "time")
stopifnot(nrow(iv_tab) == nrow(fig3))
iv_tab |>
  rename(`Time (h)` = time) |>
  knitr::kable(digits = 2, caption = "Simulated IV percentiles (with residual error) vs Figure 3.")
```

The simulated medians fall inside the model-predicted median bands drawn in
Figure 3 (about 0.35-1.5 ng/mL at -24 h, 0.45-0.65 ng/mL at 12 h) and sit
slightly below the observed medians, and roughly half to two-thirds of the
simulated 24-h values are below the 0.2 ng/mL limit of quantification, as in
the figure's lower panel. Figure 3 itself shows the model somewhat
overpredicting the spread of the data, which the paper notes (Sect. 3.2).

# Oral model

In study 92NNBT005 (microtablets chewed, repeated doses) absorption is
first-order directly into `central` with F = 39.5%; in 93NNBT007 (tablets
swallowed whole, single dose) the drug passes through three transit
compartments and F is 22.3%. The typical single-dose profiles after 2 mg
follow.

```{r oral-typical}
oral_typ <- bind_rows(lapply(0:1, function(st) {
  s <- solve_typical(
    mods$oral,
    tibble(time = 0, cmt = "depot", amt = 2, rate = NA_real_),
    list(WT = 70, STUDY_93NNBT007 = st)
  )
  tibble(time = s$time, Cc = s$Cc, study = if (st == 1) "93NNBT007" else "92NNBT005")
}))
oral_typ |>
  filter(time <= 12) |>
  ggplot(aes(time, Cc, colour = study)) +
  geom_line() +
  labs(
    x = "Time (h)", y = "Nicotine (ng/mL)", colour = "Study",
    title = "Oral nicotine 2 mg, typical 70-kg subject",
    caption = "Structure of Olsson Gisleskog 2021 Sect. 3.3 and Table 4."
  )
```

```{r oral-mass-balance}
# Study 93NNBT007 has no time effect; for 92NNBT005 the transient clearance
# increase is switched off so that CL is constant and the identity applies.
oral_off <- rxode2::ini(mods$oral, lcl_time_max = -50)
for (st in 0:1) {
  s <- solve_typical(
    oral_off,
    tibble(time = 0, cmt = "depot", amt = 2, rate = NA_real_),
    list(WT = 70, STUDY_93NNBT007 = st)
  )
  f_oral <- exp(th(mods$oral, if (st == 1) "lfdepot_93nnbt007" else "lfdepot_92nnbt005"))
  mass_balance[[paste0("oral_", if (st == 1) "93NNBT007" else "92NNBT005")]] <- mb_error(s, 2 * f_oral)
}
```

## Replicating Figure 7: time-dependent clearance and bioavailability

Repeated oral and buccal dosing gave lower-than-expected accumulation, which
the paper describes with a transient increase in clearance (oral, mouth
spray, lozenge, inhaler) or a decrease in oral bioavailability (chewing gum)
following Eq. 2, where T is the time since the first dose of the treatment
period. The typical time courses:

```{r fig7}
eq2 <- function(t, start, dur, pow) {
  ifelse(t > 0, 1 / (1 + (start / t)^pow) - 1 / (1 + ((start + dur) / t)^pow), 0)
}
tt <- seq(0, 24, by = 0.05)
typ_eff <- function(k) {
  m <- mods[[k]]
  if (k == "gum") {
    emax <- -plogis(th(m, "logitfdepot_oral_time_max"))
    start <- exp(th(m, "lfdepot_oral_t50"))
    dur <- exp(th(m, "lfdepot_oral_time_dur"))
    pow <- th(m, "fdepot_oral_time_hill")
  } else {
    emax <- exp(th(m, "lcl_time_max"))
    start <- exp(th(m, "lcl_t50"))
    dur <- exp(th(m, "lcl_time_dur"))
    pow <- exp(th(m, "lcl_time_hill"))
  }
  tibble(time = tt, change = 100 * emax * eq2(tt, start, dur, pow), model = k)
}
fig7 <- bind_rows(lapply(c("oral", "mouthspray", "lozenge", "inhaler", "gum"), typ_eff))
ggplot(fig7, aes(time, change, colour = model)) +
  geom_line() +
  labs(
    x = "Time since the first dose (h)", y = "Change (%)", colour = NULL,
    title = "Typical change in CL (oral, spray, lozenge, inhaler) or oral F (gum)",
    caption = "Replicates the typical-value lines of Figure 7 of Olsson Gisleskog 2021."
  )
fig7_max <- fig7 |>
  group_by(model) |>
  summarise(change = change[which.max(abs(change))], .groups = "drop")
stopifnot(nrow(fig7_max) == 5)
fig7_max |>
  rename(Model = model, `Largest typical change (%)` = change) |>
  knitr::kable(digits = 1)
```

The paper summarises the buccal clearance increases as "around 30%"; the
typical maxima above are
`r paste(round(range(fig7_max$change[fig7_max$model %in% c("mouthspray", "lozenge", "inhaler")])), collapse = "-")`%
for the mouth spray, lozenge and inhaler (the mouth-spray and lozenge Emax
estimates of 36% and 54% are not reached because onset and offset overlap),
with the individual
curves in Figure 7 varying widely around them because the onset, duration or
Emax carry IIV. The oral repeated-dose study and the gum have larger, shorter
or longer-lasting changes, as in Figure 7.

# Buccal models

## Fraction swallowed

```{r frsw}
fsw_tab <- tibble(
  Product = c("Mouth spray 2 mg", "Mouth spray 4 mg", "Chewing gum", "Lozenge (NiQuitin) 2 mg", "Lozenge (NiQuitin) 4 mg", "Inhaler"),
  Model = c(
    100 * plogis(th(mods$mouthspray, "logitfsw")) * (c(2, 4) / 2)^th(mods$mouthspray, "e_dose_nicotine_mg_fsw"),
    100 * plogis(th(mods$gum, "logitfsw")),
    100 * plogis(th(mods$lozenge, "logitfsw") + c(th(mods$lozenge, "e_dose2mg_fsw"), 0)),
    100 * plogis(th(mods$inhaler, "logitfsw"))
  ),
  `Table 5` = c(60.6, 64.6, 54.7, 64.0, 68.8, 66.9)
)
knitr::kable(fsw_tab, digits = 1, caption = "Typical fraction of the dose swallowed (%).")
stopifnot(all(abs(fsw_tab$Model - fsw_tab$`Table 5`) < 0.051))
```

These are also the values quoted in the Discussion: lowest for chewing gum
(55%), then mouth spray (61%), inhaler (67%) and lozenge (69%). A Nicorette
lozenge adds 0.412 on the logit scale, giving
`r round(100 * plogis(th(mods$lozenge, "logitfsw") + th(mods$lozenge, "e_form_nicotine_nicorette_lozenge_fsw")), 1)`%
at 4 mg.

## Mass balance

With the time-dependent change switched off, `CL * AUC(0-inf)` must equal the
dose times `(1 - Frsw) * F_buccal + Frsw * F_oral`. For the gum only the
amount released during the 0.5-h chew counts: `1 - exp(-Krel (0.5 - lag))` of
the buccal share, whose release starts after the lag, and
`1 - exp(-Krel 0.5)` (= released/nominal) of the swallowed share.

```{r buccal-mass-balance}
D <- 2
f_or <- 0.395

m <- rxode2::ini(mods$mouthspray, lcl_time_max = -50)
fsw <- plogis(th(m, "logitfsw"))
s <- solve_typical(m, tibble(time = 0, cmt = c("depot_buccal", "depot_oral"), amt = D, rate = NA_real_),
  list(WT = 70, OCC = 1, DOSE_NICOTINE_MG = D, ROUTE_SUBLINGUAL = 0))
mass_balance$mouthspray <- mb_error(s, D * ((1 - fsw) + fsw * f_or))

released_gum <- D * (1 - exp(-2.8 * 0.5)) # the paper's average Krel of 2.8 1/h
krel_gum <- -log(1 - released_gum / D) / 0.5
lag_gum <- exp(th(mods$gum, "ltlag"))
fsw <- plogis(th(mods$gum, "logitfsw"))
s <- solve_typical(mods$gum, tibble(time = 0, cmt = c("depot", "depot_oral"), amt = D, rate = NA_real_),
  list(WT = 70, OCC = 1, DOSE_NICOTINE_MG = D, DOSE_NICOTINE_RELEASED_MG = released_gum, FORM_NICOTINE_FRESHMINT = 0))
mass_balance$gum <- mb_error(s, D * ((1 - fsw) * (1 - exp(-krel_gum * (0.5 - lag_gum))) + fsw * f_or * (1 - exp(-krel_gum * 0.5))))

m <- rxode2::ini(mods$lozenge, lcl_time_max = -50)
fsw <- plogis(th(m, "logitfsw") + th(m, "e_form_nicotine_nicorette_lozenge_fsw") + th(m, "e_dose2mg_fsw"))
s <- solve_typical(m, tibble(time = 0, cmt = c("depot", "depot_oral"), amt = D, rate = NA_real_),
  list(WT = 70, OCC = 1, DOSE_NICOTINE_MG = D, FORM_NICOTINE_NICORETTE_LOZENGE = 1))
mass_balance$lozenge <- mb_error(s, D * ((1 - fsw) + fsw * f_or))

m <- rxode2::ini(mods$inhaler, lcl_time_max = -50)
fsw <- plogis(th(m, "logitfsw"))
for (st in 0:1) {
  s <- solve_typical(m, tibble(time = 0, cmt = c("depot_buccal", "depot_oral"), amt = D, rate = D * 3),
    list(WT = 70, OCC = 1, STUDY_97NNIN024 = st))
  f_study <- 1 + th(m, "e_study_97nnin024_f") * st
  mass_balance[[paste0("inhaler_study97NNIN024_", st)]] <- mb_error(s, D * f_study * ((1 - fsw) + fsw * f_or))
}
```

## Replicating Figure 11

Figure 11 compares simulated single doses of a 2-mg mouth spray, 2-mg
Nicorette classic gum, 2-mg Nicorette lozenge and the inhaler (means and 90%
prediction intervals). The gum is simulated with the released amount that
gives the paper's average release rate of 2.8 1/h (75% of the nominal dose),
and the inhaler with 2 mg released over a 20-min session, the average reported
in Sect. 3.4.

```{r fig11}
t_buc <- c(seq(0, 1, by = 1 / 12), seq(1.25, 3, by = 0.25), 3.5, 4:12)
ev_buc <- bind_rows(
  make_arm(tibble(time = 0, cmt = c("depot_buccal", "depot_oral"), amt = 2, rate = NA_real_), t_buc,
    list(OCC = 1, DOSE_NICOTINE_MG = 2, ROUTE_SUBLINGUAL = 0), "Mouth spray 2 mg", 0L),
  make_arm(tibble(time = 0, cmt = c("depot", "depot_oral"), amt = 2, rate = NA_real_), t_buc,
    list(OCC = 1, DOSE_NICOTINE_MG = 2, DOSE_NICOTINE_RELEASED_MG = released_gum, FORM_NICOTINE_FRESHMINT = 0),
    "Nicorette gum classic 2 mg", 1000L),
  make_arm(tibble(time = 0, cmt = c("depot", "depot_oral"), amt = 2, rate = NA_real_), t_buc,
    list(OCC = 1, DOSE_NICOTINE_MG = 2, FORM_NICOTINE_NICORETTE_LOZENGE = 1), "Nicorette lozenge 2 mg", 2000L),
  make_arm(tibble(time = 0, cmt = c("depot_buccal", "depot_oral"), amt = 2, rate = 2 * 3), t_buc,
    list(OCC = 1, STUDY_97NNIN024 = 0), "Nicorette inhaler (2 mg released)", 3000L)
)
stopifnot(!anyDuplicated(unique(ev_buc[, c("id", "time", "evid", "cmt")])))
buc_model <- c(
  "Mouth spray 2 mg" = "mouthspray", "Nicorette gum classic 2 mg" = "gum",
  "Nicorette lozenge 2 mg" = "lozenge", "Nicorette inhaler (2 mg released)" = "inhaler"
)
sim_buc <- bind_rows(lapply(names(buc_model), function(trt) {
  ev <- ev_buc |>
    filter(treatment == trt) |>
    select(where(~ !all(is.na(.x))))
  rxode2::rxSolve(mods[[buc_model[[trt]]]], ev, keep = "treatment", returnType = "data.frame")
}))
buc_sum <- sim_buc |>
  group_by(treatment, time) |>
  summarise(mean = mean(Cc), p05 = quantile(Cc, 0.05), p95 = quantile(Cc, 0.95), .groups = "drop")
ggplot(buc_sum, aes(time, mean, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.15, colour = NA) +
  geom_line() +
  labs(
    x = "Time (h)", y = "Nicotine (ng/mL)", colour = NULL, fill = NULL,
    title = "Single doses of four buccal products",
    caption = "Replicates Figure 11 of Olsson Gisleskog 2021 (means and 90% prediction intervals)."
  ) +
  guides(colour = guide_legend(ncol = 2), fill = guide_legend(ncol = 2))
```

```{r fig11-compare}
# Peaks of the MEAN curves read by eye from the raster Figure 11 (about
# +/- 0.2 ng/mL, +/- 5 min). The published inhaler curve is labelled
# '15-mg inhaler' without the released amount, so it is shown but not gated.
fig11 <- tibble(
  treatment = names(buc_model),
  `Figure 11 peak (ng/mL)` = c(4.9, 3.5, 3.6, 4.5),
  `Figure 11 time of peak (h)` = c(0.17, 0.5, 1.0, 1.0),
  gated = c(TRUE, TRUE, TRUE, FALSE)
)
fig11_cmp <- buc_sum |>
  group_by(treatment) |>
  summarise(`Simulated peak (ng/mL)` = max(mean), `Simulated time of peak (h)` = time[which.max(mean)], .groups = "drop") |>
  inner_join(fig11, by = "treatment") |>
  mutate(`Difference (%)` = 100 * (`Simulated peak (ng/mL)` / `Figure 11 peak (ng/mL)` - 1))
stopifnot(nrow(fig11_cmp) == 4)
fig11_cmp |>
  select(-gated) |>
  knitr::kable(digits = 2, caption = "Peak of the simulated mean curve vs Figure 11.")
# Over 8 seeds (identical at 1 and 4 solver threads) the differences were
# -3.5 to +4.2% (spray), +7.8 to +18.8% (gum; systematically high, see the
# text) and -7.5 to +2.7% (lozenge), so 30% leaves at least 11 points of
# headroom. A transcription error in a rate constant, a fraction swallowed or
# a unit moves the peak by far more.
stopifnot(all(abs(fig11_cmp$`Difference (%)`[fig11_cmp$gated]) < 30))
```

The mouth spray, gum and lozenge reproduce Figure 11 in both height and
timing of the mean peak. The gum peak is about 10% higher than the figure in
every cohort drawn while developing this article, which is consistent with the
released amount assumed here (the paper does not state the one it simulated).
The simulated inhaler peak is lower than the published one (see Assumptions
and deviations).

# Transdermal model

## Mass balance

The first-order pathway is scaled so that the whole fraction Fr1 of the
bioavailable dose leaves the patch within its release window of Frdur1 x 16 h.
For the Invisipatch the window is timed from application but the first-order
release only starts after the 0.53-h lag, so slightly less than Fr1 is
released; the zero-order pathway delivers `FTOT (1 - Fr1)` between 4.06 h and
patch removal.

```{r td-mass-balance}
m <- rxode2::ini(mods$transdermal, lcl_time_max = -50)
for (inv in 0:1) {
  sfx <- if (inv == 1) "invisipatch" else "nicorette"
  s <- solve_typical(m, tibble(time = 0, cmt = c("depot_td", "transit1"), amt = 25, rate = c(NA_real_, -2)),
    list(WT = 70, OCC = 1, FORM_NICOTINE_INVISIPATCH = inv, T_PATCH_WEAR = 16))
  ftot <- exp(th(m, "lfdepot"))
  ffo <- plogis(th(m, paste0("logitffo_", sfx)))
  dd <- 16 * plogis(th(m, paste0("logitfrdur_", sfx)))
  kr <- exp(th(m, "lkrel"))
  lag1 <- if (inv == 1) exp(th(m, "ltlag_invisipatch")) else 0
  expected <- 25 * ftot * (ffo * (1 - exp(-kr * (dd - lag1))) / (1 - exp(-kr * dd)) + (1 - ffo))
  mass_balance[[paste0("patch_", sfx)]] <- mb_error(s, expected)
}
```

## Replicating Figure 12

```{r fig12}
t_td <- c(seq(0, 2, by = 0.5), 3:24)
ev_td <- bind_rows(
  make_arm(tibble(time = 0, cmt = c("depot_td", "transit1"), amt = 25, rate = c(NA_real_, -2)), t_td,
    list(OCC = 1, FORM_NICOTINE_INVISIPATCH = 0, T_PATCH_WEAR = 16), "Nicorette 25 mg/16 h", 0L),
  make_arm(tibble(time = 0, cmt = c("depot_td", "transit1"), amt = 25, rate = c(NA_real_, -2)), t_td,
    list(OCC = 1, FORM_NICOTINE_INVISIPATCH = 1, T_PATCH_WEAR = 16), "Invisipatch 25 mg/16 h", 1000L)
)
stopifnot(!anyDuplicated(unique(ev_td[, c("id", "time", "evid", "cmt")])))
sim_td <- rxode2::rxSolve(mods$transdermal, ev_td, keep = "treatment", returnType = "data.frame")
td_sum <- sim_td |>
  group_by(treatment, time) |>
  summarise(mean = mean(Cc), p05 = quantile(Cc, 0.05), p95 = quantile(Cc, 0.95), .groups = "drop")
ggplot(td_sum, aes(time, mean, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.15, colour = NA) +
  geom_line() +
  scale_x_continuous(breaks = seq(0, 24, 4)) +
  labs(
    x = "Time (h)", y = "Nicotine (ng/mL)", colour = NULL, fill = NULL,
    title = "Single 25-mg/16-h patch applications",
    caption = "Replicates Figure 12 of Olsson Gisleskog 2021 (means and 90% prediction intervals)."
  )
```

```{r fig12-compare}
# Mean curves read by eye from the raster Figure 12 (about +/- 0.5 ng/mL).
fig12 <- tribble(
  ~treatment, ~time, ~figure12,
  "Nicorette 25 mg/16 h", 4, 17.0,
  "Nicorette 25 mg/16 h", 8, 16.2,
  "Nicorette 25 mg/16 h", 12, 13.0,
  "Nicorette 25 mg/16 h", 16, 11.7,
  "Nicorette 25 mg/16 h", 24, 2.6,
  "Invisipatch 25 mg/16 h", 4, 13.6,
  "Invisipatch 25 mg/16 h", 8, 15.8,
  "Invisipatch 25 mg/16 h", 12, 14.2,
  "Invisipatch 25 mg/16 h", 16, 11.7,
  "Invisipatch 25 mg/16 h", 24, 2.6
)
fig12_cmp <- fig12 |>
  inner_join(td_sum |> select(treatment, time, simulated = mean), by = c("treatment", "time")) |>
  mutate(`Difference (%)` = 100 * (simulated / figure12 - 1))
stopifnot(nrow(fig12_cmp) == nrow(fig12))
fig12_cmp |>
  rename(`Time (h)` = time, `Figure 12 mean (ng/mL)` = figure12, `Simulated mean (ng/mL)` = simulated) |>
  knitr::kable(digits = 1, caption = "Simulated mean concentration vs Figure 12.")
# Invisipatch during wear (4-16 h): gated. Over 8 seeds the differences were
# -10.9 to +8.5%, so 25% leaves 14 points of headroom. The 24-h point (8 h
# after removal) and the Nicorette patch are discussed below and not gated.
inv_wear <- fig12_cmp |> filter(treatment == "Invisipatch 25 mg/16 h", time <= 16)
stopifnot(nrow(inv_wear) == 4, all(abs(inv_wear$`Difference (%)`) < 25))
```

The Invisipatch reproduces Figure 12 during the application period. The
Nicorette patch does not: the model rises more slowly
(`r round(fig12_cmp$"Difference (%)"[fig12_cmp$treatment == "Nicorette 25 mg/16 h" & fig12_cmp$time == 4])`%
against the published mean at 4 h) and stays higher towards removal
(`r sprintf("%+.0f", fig12_cmp$"Difference (%)"[fig12_cmp$treatment == "Nicorette 25 mg/16 h" & fig12_cmp$time == 16])`%
at 16 h). The simulated
Nicorette profile does, however, agree with the simulated median of the
paper's own VPC for the 25-mg Nicorette patch (Figure 10, STRT 3: about
12-14 ng/mL from 8 to 12 h, peak observed around 10 h, Sect. 3.5), and the
parameters are those printed in Table 6 and in the final NONMEM output. The
difference is therefore attributed to Figure 12 rather than to the model; see
Assumptions and deviations.

## Replicating Figure 10 (STRT 7 and 8): three daily Invisipatch applications

```{r fig10}
t_rep <- c(seq(0, 72, by = 1))
ev_rep <- bind_rows(lapply(c(16, 24), function(wear) {
  make_arm(
    tibble(time = rep(c(0, 24, 48), each = 2), cmt = rep(c("depot_td", "transit1"), 3), amt = 25, rate = rep(c(NA_real_, -2), 3)),
    t_rep, list(OCC = 1, FORM_NICOTINE_INVISIPATCH = 1, T_PATCH_WEAR = wear),
    paste0("Invisipatch 25 mg worn ", wear, " h"), if (wear == 16) 0L else 1000L
  )
}))
sim_rep <- rxode2::rxSolve(mods$transdermal, ev_rep, keep = "treatment", returnType = "data.frame")
sim_rep |>
  group_by(treatment, time) |>
  summarise(p05 = quantile(Cc, 0.05), p50 = median(Cc), p95 = quantile(Cc, 0.95), .groups = "drop") |>
  ggplot(aes(time, p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.2) +
  geom_line() +
  facet_wrap(~treatment) +
  scale_x_continuous(breaks = seq(0, 72, 12)) +
  labs(
    x = "Time (h)", y = "Nicotine (ng/mL)",
    title = "Three daily 25-mg Invisipatch applications",
    caption = "Replicates the simulated median and 5th-95th percentiles of Figure 10\n(STRT 7 and 8) of Olsson Gisleskog 2021."
  )
```

```{r fig10-table}
rep_tab <- sim_rep |>
  group_by(treatment, time) |>
  summarise(median = median(Cc), .groups = "drop") |>
  group_by(treatment) |>
  summarise(
    `Day 1 peak (ng/mL)` = max(median[time <= 24]),
    `Trough before day 2 (ng/mL)` = median[time == 24],
    `Day 3 peak (ng/mL)` = max(median[time >= 48 & time <= 72]),
    .groups = "drop"
  )
stopifnot(nrow(rep_tab) == 2)
knitr::kable(rep_tab, digits = 1, caption = "Median profile features for three daily 25-mg Invisipatch applications.")
```

Sect. 3.5 reports that keeping the patch on for 24 h raised the trough
without a large difference in peak exposure; the table shows the same pattern
in the simulation.

# Mass-balance summary

```{r mass-balance-summary}
mb <- tibble(Check = names(mass_balance), `Relative error` = unlist(mass_balance))
knitr::kable(mb, digits = 8, caption = "CL * AUC(0-inf) against the expected bioavailable amount (typical values, tight solver tolerances).")
# Realised errors are below 1e-5 (trapezoid rule on the dense grid); a
# mis-set bioavailability, dose split, lag or release window moves them by
# at least a few percent.
stopifnot(length(mass_balance) == 10, all(abs(mb$`Relative error`) < 1e-4))
```

# PKNCA

The paper reports no non-compartmental results. For reference, PKNCA
summaries of the simulated single-dose cohorts of Figures 11 and 12 follow.

```{r pknca}
nca_one <- function(sim, end) {
  conc <- sim |>
    filter(!is.na(Cc)) |>
    mutate(Cc = pmax(Cc, 0)) |>
    select(id, time, Cc, treatment)
  conc <- bind_rows(conc, distinct(conc, id, treatment) |> mutate(time = 0, Cc = 0)) |>
    distinct(id, treatment, time, .keep_all = TRUE) |>
    arrange(treatment, id, time)
  dose <- distinct(conc, id, treatment) |> mutate(time = 0, amt = 1)
  conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ time | treatment + id, concu = "ng/mL", timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(dose, amt ~ time | treatment + id, doseu = "mg")
  intervals <- data.frame(start = 0, end = end, cmax = TRUE, tmax = TRUE, auclast = TRUE)
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}
nca_buc <- nca_one(sim_buc, 12)
nca_td <- nca_one(sim_td, 24)
nca_tab <- bind_rows(as.data.frame(nca_buc), as.data.frame(nca_td)) |>
  group_by(treatment, PPTESTCD) |>
  summarise(median = median(PPORRES), .groups = "drop") |>
  pivot_wider(names_from = PPTESTCD, values_from = median)
stopifnot(nrow(nca_tab) == 6)
nca_tab |>
  select(treatment, cmax, tmax, auclast) |>
  rename(
    Treatment = treatment, `Cmax (ng/mL)` = cmax, `Tmax (h)` = tmax,
    `AUC0-last (ng*h/mL; 0-12 h buccal, 0-24 h patch)` = auclast
  ) |>
  knitr::kable(digits = 2, caption = "Median simulated single-dose NCA (virtual subjects without residual pre-study nicotine).")
```

# Assumptions and deviations

- **Source of the values.** Parameter values were taken from the paper's
  Tables 3-6 and cross-checked against the final NONMEM control streams and
  output listings in the electronic supplementary material. Where the paper
  prints three significant figures and the extravascular control streams
  carry the intravenous estimates to more (for example CL 67.4 vs
  67.4136 L/h), the longer value is used; the two agree to the printed
  precision.
- **Lozenge duration and pow (Table 5 misprint).** Table 5 prints 20.2 h and
  9.88 as the lozenge 'Duration' and 'pow'. The final NONMEM output gives
  7.18 h (standard error 1.45) and 3.18 (standard error 0.314), whose relative
  standard errors are 20.2% and 9.87%: the table printed each RSE in the
  estimate column. The model uses 7.18 h and 3.18. Every other lozenge value
  in Table 5 matches the output.
- **Gum ka footnotes.** Table 5 footnote f gives `Ka = Ka2mg + (1/NDOSE)^0.507`
  and footnote d refers to '3 mg of Nicorette classic'. The control stream
  computes `KA = THETA * (NDOSE/2)^0.507` with 2 mg as the reference for both
  gums; the model follows the control stream.
- **Time since the first dose.** Eq. 2 uses the time since the first dose of
  the study period (dataset column TSSP). The models compute it with
  rxode2's `tafd()` on a compartment that has no lag (the gut dose for the
  buccal products; the patch dose plus its lag for the patch). NONMEM
  evaluated TSSP only at data records, so its clearance changed in steps
  between records while rxode2 changes it continuously. For multi-period
  (crossover) simulations, simulate each period as its own subject or reset
  the system between periods, and set `OCC` to the period number: `tafd()`
  counts from the first dose ever given to the subject.
- **Inhaler input.** The inhaler dose records carry a fixed infusion rate
  (released amount over 20 min), as in the source data. With a fixed rate,
  bioavailability shortens the input rather than lowering its rate (NONMEM
  and rxode2 behave the same way), so the effective input lasts
  `F x 20 min`. Supply `rate = amt * 3` rather than a modelled duration.
- **Parameters fixed to zero in the source.** The inhaler control stream
  carries an IIV on the onset time of the clearance change fixed to 0, and the
  lozenge control stream a NiQuitin effect on Krel fixed to 0. Both are
  omitted: they have no effect, and a zero variance would make the OMEGA
  matrix singular.
- **Patch release window after a repeat application.** The first-order
  window is timed from each application. For the Invisipatch, rxode2 measures
  the time from the lagged (0.53 h) arrival of the new dose, so during the
  first 0.53 h after a repeat application the small residue left in the
  first-order pool from the previous patch is not released as it would be in
  NONMEM. The effect on concentrations is negligible.
- **Figure 12, Nicorette patch.** The simulated mean for the 25-mg Nicorette
  patch is about 40% below Figure 12 at 4 h and 10-15% above it at 16 h,
  while the Invisipatch agrees. The model uses the published Table 6 values,
  which match the final NONMEM output, and it agrees with the paper's own VPC
  for this product (Figure 10, STRT 3). No correction notice for the article
  was found (checked 2026-09-28). The difference is left as a documented
  deviation; the parameters were not adjusted.
- **Figure 11, inhaler.** The published curve is labelled '15-mg inhaler'
  without the amount released or the number of sessions. Simulating the
  reported average of 2 mg released in one 20-min session gives a mean peak
  of
  `r round(fig11_cmp$"Simulated peak (ng/mL)"[fig11_cmp$treatment == "Nicorette inhaler (2 mg released)"], 1)`
  ng/mL against about 4.5 ng/mL in the figure. The inhaler is therefore not
  gated against Figure 11.
- **Simulation inputs not given in the paper.** Body weights are drawn from
  a log-normal distribution matching Table 2. The gum simulation uses a
  released amount of 75% of the nominal dose, which reproduces the paper's
  average release rate constant of 2.8 1/h. The simulations for Figures 11
  and 12 omit residual pre-study nicotine (no virtual pre-washout dose).
- **Data below the limit of quantification.** The intravenous model was fitted
  with the M3 method and the other models excluded such data. This affects
  estimation only and has no counterpart in the simulation models.
