---
title: "Meropenem (Onichimowski 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Meropenem (Onichimowski 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Onichimowski_2020_meropenem"))$reference`
* Description: `r rxode2::rxode(readModelDb("Onichimowski_2020_meropenem"))$description`
* Article: <https://doi.org/10.1007/s43440-020-00104-3> (open access)

# Population

Onichimowski et al. ran a prospective, single-centre observational study in a
tertiary medical/surgical ICU in Olsztyn, Poland. Twenty critically ill adults
on continuous renal replacement therapy (CRRT) were enrolled and 19 contributed
PK data (5 female, 14 male; age 36-79 years, median 67; body weight 60-100 kg,
median 80). Nine patients were on CVVH with heparin anticoagulation and ten on
CVVHD with regional citrate anticoagulation, all with an AV 1000 polysulfone
1.8 m^2 filter. Ten were septic. Median serum albumin was 24.6 g/L (range
15.6-31.8), median APACHE II 31 and median SOFA 10 (Table 1). Every patient
received meropenem 1 g as a 1-h infusion every 8 h. Arterial samples were taken
before the dose and at 15, 30, 45, 60, 75, 90, 120, 180, 240 and 480 min after
the start of the infusion, giving 256 concentrations.

# Source trace

| Element | Value | Source location |
|---|---|---|
| Structure | Two compartments, IV infusion, elimination from central (ClCRRT) | Methods Eq. 1-2; ADVAN3 TRANS4 |
| `lvc` | V1 = 27.9 L at ALB = 24.6 g/L | Table 2 |
| `lcl` | ClCRRT = 15.1 L/h | Table 2 |
| `lq` | Q = 21.1 L/h | Table 2 |
| `lvp` | V2 = 33.7 L | Table 2 |
| `e_alb_vc` | -2.87 (power on ALB / 24.6) | Table 2; Results equation `V1,i = 27.9 (ALB_i/24.6)^-2.87 exp(eta)` |
| `etalvc` | 53.1 %CV -> omega^2 = log(0.531^2 + 1) = 0.248 | Table 2 and footnote |
| `etalcl` | 43.7 %CV -> 0.175 | Table 2 and footnote |
| `etalvp` | 85.6 %CV -> 0.550 | Table 2 and footnote |
| IIV on Q | 0, fixed (omitted) | Table 2; Results text |
| `addSd` | 0.881 mg/L | Table 2 |
| `propSd` | 24.1 %CV -> 0.241 | Table 2; Methods Eq. 4 |

# Model

```{r model}
mod <- readModelDb("Onichimowski_2020_meropenem")
mod_ui <- rxode2::rxode(mod)
mod_typ <- rxode2::zeroRe(mod_ui)
```

## V1 against albumin (Figure 2, upper panel)

```{r fig2}
alb_grid <- data.frame(ALB = seq(15, 32, by = 0.5)) |>
  mutate(V1 = 27.9 * (ALB / 24.6)^-2.87)
ggplot(alb_grid, aes(ALB, V1)) +
  geom_line() +
  labs(x = "Albumin (g/L)", y = "Typical V1 (L)",
       title = "Replicates Figure 2 (upper panel) of Onichimowski 2020")
```

## Steady-state volume

Table 3 reports Vss = V1 + V2 = 61.6 L for a patient with albumin 24.6 g/L. The
typical values in the packaged model reproduce this exactly.

```{r vss}
ini_df <- mod_ui$iniDf
th <- setNames(ini_df$est, ini_df$name)
vss <- exp(th[["lvc"]]) + exp(th[["lvp"]])
vss
stopifnot(abs(vss - 61.6) < 0.05)
```

# Typical first-dose and steady-state profiles

```{r typical}
alb_levels <- c(15.6, 24.6, 31.8)
ev_typ <- rxode2::et(amt = 1000, dur = 1, ii = 8, addl = 29, cmt = "central") |>
  rxode2::et(c(seq(0, 8, by = 0.05), seq(232, 240, by = 0.05)), cmt = "central") |>
  as.data.frame()
typ <- bind_rows(lapply(alb_levels, function(a) {
  d <- ev_typ
  d$ALB <- a
  s <- as.data.frame(rxode2::rxSolve(mod_typ, d, returnType = "data.frame"))
  s$ALB <- a
  s
}))
typ |>
  mutate(
    period = ifelse(time <= 8, "First dose", "Steady state (dose 30)"),
    tad = ifelse(time <= 8, time, time - 232)
  ) |>
  ggplot(aes(tad, Cc, colour = factor(ALB))) +
  geom_line() +
  facet_wrap(~period) +
  scale_y_log10() +
  labs(x = "Time after start of infusion (h)", y = "Meropenem (mg/L)",
       colour = "Albumin (g/L)",
       title = "Typical profiles, 1 g as a 1-h infusion q8h")
```

The observed first-dose concentrations in Figure 1 of the paper peak at roughly
20-60 mg/L and fall to about 2-10 mg/L at 8 h, consistent with the typical
curves above.

# Probability of target attainment (Figure 4)

The paper simulated 1000 subjects at steady state after 1 g as a 1-h infusion
q8h for albumin 15.6, 24.6 and 31.8 g/L and counted the fraction reaching
40% and 100% T > MIC over 24 h. Here 200 virtual subjects per albumin level
are simulated with between-subject variability on the final estimates; the
last 24 h of a 30-dose course is used as steady state.

```{r pta-sim}
rxode2::rxSetSeed(20200416)
n_per_arm <- 200
obs_ss <- seq(216, 240, by = 0.1)
ev_pta <- bind_rows(lapply(seq_along(alb_levels), function(i) {
  e <- rxode2::et(amt = 1000, dur = 1, ii = 8, addl = 29, cmt = "central") |>
    rxode2::et(obs_ss, cmt = "central") |>
    rxode2::et(id = seq_len(n_per_arm) + (i - 1) * n_per_arm) |>
    as.data.frame()
  e$ALB <- alb_levels[i]
  e
}))
sim_pta <- as.data.frame(rxode2::rxSolve(mod_ui, ev_pta, keep = "ALB",
                                         returnType = "data.frame"))
```

```{r pta-calc}
ft_mic <- function(sim, mic) {
  sim |>
    filter(time >= 216) |>
    group_by(ALB, id) |>
    summarise(ft = mean(Cc > mic), .groups = "drop")
}
mics <- 2^seq(-4, 6, by = 0.5)
pta <- bind_rows(lapply(mics, function(m) {
  ft_mic(sim_pta, m) |>
    group_by(ALB) |>
    summarise(pta40 = mean(ft >= 0.4), pta100 = mean(ft >= 1), .groups = "drop") |>
    mutate(MIC = m)
}))
pta |>
  pivot_longer(c(pta40, pta100), names_to = "target", values_to = "PTA") |>
  mutate(target = ifelse(target == "pta40", "40% T > MIC", "100% T > MIC")) |>
  ggplot(aes(MIC, PTA * 100, colour = factor(ALB))) +
  geom_line() +
  geom_hline(yintercept = 90, linetype = "dashed") +
  scale_x_log10() +
  facet_wrap(~target) +
  labs(x = "MIC (mg/L)", y = "PTA (%)", colour = "Albumin (g/L)",
       title = "Replicates Figure 4a and 4c of Onichimowski 2020")
```

```{r pta-table}
pta2 <- ft_mic(sim_pta, 2) |>
  group_by(ALB) |>
  summarise(pta40 = 100 * mean(ft >= 0.4), pta100 = 100 * mean(ft >= 1), .groups = "drop") |>
  mutate(
    pub40 = c(99.8, 97.8, 95.3),
    pub100 = c(91.8, 70.0, 58.7)
  )
pta2 |>
  select(ALB, pta40, pub40, pta100, pub100) |>
  dplyr::rename(
    "Albumin (g/L)" = ALB,
    "Simulated PTA 40% T>2 mg/L (%)" = pta40,
    "Published PTA 40% T>2 mg/L (%)" = pub40,
    "Simulated PTA 100% T>2 mg/L (%)" = pta100,
    "Published PTA 100% T>2 mg/L (%)" = pub100
  ) |>
  knitr::kable(digits = 1)
```

The 40% target is reached by nearly every subject at all three albumin levels,
and the 100% target falls with rising albumin, as in the paper. The published
PTA values are bootstrap medians that fold parameter uncertainty in on top of
between-subject variability, and each simulated arm here has 200 subjects
(Monte Carlo SE about 3 percentage points at 70%), so agreement is checked
with a band rather than exactly.

```{r pta-check}
stopifnot(
  all(pta2$pta40 > 88),
  # ordering in albumin is the covariate effect the paper reports
  pta2$pta100[1] > pta2$pta100[3],
  all(abs(pta2$pta100 - pta2$pub100) < 15)
)
```

# PKNCA validation

The paper reports no NCA. As an internal check, PKNCA is run on the typical
(no-variability) steady-state interval at each albumin level; at steady state
AUC over one dosing interval must equal dose / CL = 1000 / 15.1 = 66.2 mg*h/L
regardless of albumin, because albumin acts only on V1.

```{r pknca}
nca_in <- typ |>
  filter(time >= 232, !is.na(Cc)) |>
  mutate(id = match(ALB, alb_levels), treatment = paste0("ALB ", ALB, " g/L"))
dose_df <- nca_in |>
  distinct(id, treatment) |>
  mutate(time = 232, amt = 1000)
conc_obj <- PKNCA::PKNCAconc(nca_in, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)
intervals <- data.frame(start = 232, end = 240, cmax = TRUE, tmax = TRUE,
                        cmin = TRUE, auclast = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_tab <- as.data.frame(nca_res) |>
  select(treatment, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
nca_tab |>
  dplyr::rename(
    "Group" = treatment,
    "Cmax (mg/L)" = cmax,
    "Tmax (h)" = tmax,
    "Cmin (mg/L)" = cmin,
    "AUC0-8,ss (mg*h/L)" = auclast
  ) |>
  knitr::kable(digits = 2)
stopifnot(all(abs(nca_tab$auclast / (1000 / 15.1) - 1) < 0.01))
```

# Assumptions and deviations

* IIV on Q was fixed to zero by the authors because it tended to zero; it is
  omitted from `ini()` rather than carried as a zero-variance eta.
* Table 2 gives IIV as %CV with the footnote `%CV = sqrt(exp(IIV^2 - 1) * 100%`;
  the misplaced parenthesis is read as the standard log-normal conversion
  `CV = sqrt(exp(omega^2) - 1)`, so `omega^2 = log(CV^2 + 1)`.
* The proportional residual error is reported as "sigma^2 prop, %CV = 24.1" and
  is encoded as a proportional SD of 0.241; the additive term (0.881 mg/L) is
  taken as an SD.
* The model has a single elimination clearance, which the authors label ClCRRT;
  no residual renal or non-renal clearance was separated from it, and CRRT
  settings were not retained as covariates.
* The simulated PTA uses the point estimates only; the paper's values also
  integrate bootstrap parameter uncertainty, so their confidence intervals are
  wider than the between-subject variability alone produces.
