---
title: "Fentanyl (Oosten 2016)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fentanyl (Oosten 2016)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Oosten_2016_fentanyl"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Oosten_2016_fentanyl"))$description`
* Article: <https://doi.org/10.1007/s00228-015-2005-x> (open access; Eur J Clin Pharmacol 2016;72(4):459-467)

# Population

The model was developed from 942 fentanyl plasma samples collected in 52
adult cancer patients (3 of whom participated twice) admitted to the
Erasmus MC Cancer Institute (Rotterdam, The Netherlands) between January
2010 and November 2013 for moderate-to-severe cancer-related nociceptive
pain (Oosten 2016 Table 1). Median age 63 years (range 23-80); 33 male
(63%), 19 female (37%); 47 (90%) Caucasian. Median body mass index was
25 kg/m^2 (range 18-40). Primary tumor sites were breast (15%), urinary
tract including kidney (15%), prostate (13%), soft-tissue sarcoma / GIST
(12%), colorectal (10%), and other (35%).

Subcutaneous fentanyl doses ranged from 10 to 300 ug/h continuous infusion
(median 75); transdermal fentanyl Sandoz Matrix patch doses ranged from
12 to 400 ug/h (median 50), replaced every 72 h. Sampling was sparse and
opportunistic: median 15 samples per patient (range 1-86), median observed
concentration 1.33 ng/mL (range 0.122-10.7). 32 patients had semi-
simultaneous sc and td exposure; 13 had sc samples without previous td;
9 had td samples without sc; the majority (33) already used transdermal
fentanyl before admission.

The same information is available programmatically via
`readModelDb("Oosten_2016_fentanyl")$population`.

# Source trace

Per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Oosten_2016_fentanyl.R`. The
table below collects them for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Structural model | 1-cmt, two parallel first-order absorption depots, td lag time | Oosten 2016 Results, "Fentanyl pharmacokinetics" paragraph 1 |
| `lka_sc` (Ka_sc) | `log(0.0358)` 1/h | Table 2: ka_sc = 0.0358 (RSE 24.4%); bootstrap 0.0374 (95% CI 0.0248-0.0555) |
| `lka_td` (Ka_td) | `log(0.0135)` 1/h | Table 2: ka_td = 0.0135 (RSE 16.8%); bootstrap 0.0140 (95% CI 0.0105-0.0188) |
| `llag_td` (Tlag_td) | `log(4.73)` h | Table 2: t_lag_td = 4.73 (RSE 21.2%); bootstrap 4.65 (95% CI 2.25-6.98) |
| `lcl` (CL/F at 70 kg) | `log(49.6)` L/h | Table 2: CL_70kg/F = 49.6 (RSE 9.36%); bootstrap 50.4 (95% CI 40.9-61.6) |
| `lvc` (V/F at 70 kg) | `fixed(log(280))` L | Table 2 + Results: V_70kg/F fixed to 280 L (citation [25]); sensitivity-analyzed +/-50% with stable CL estimate |
| `lfdepot_td` (F_td) | `fixed(log(1))` | Table 2: typical F_td not separately reported; IIV only (anchor F_td = 1 at the population level) |
| `e_wt_cl` | `fixed(0.75)` | Methods + Results: "allometrically scaled body weight on CL/F"; canonical theoretical-allometric exponent inferred |
| `e_wt_vc` | `fixed(1)` | Methods + Results: linear weight on V/F under the same canonical-allometric inference |
| `etalka_sc` IIV Ka_sc | `log(1 + 0.935^2)` | Table 2: IIV Ka_sc 93.5% CV (RSE 15.2%); bootstrap 91.1 (95% CI 59.6-119) |
| `etalka_td` IIV Ka_td | `log(1 + 0.424^2)` | Table 2: IIV Ka_td 42.4% CV (RSE 23.9%); bootstrap 41.4 (95% CI 10.5-59.2) |
| `etalfdepot_td` IIV F_td | `log(1 + 0.423^2)` | Table 2: IIV F_td 42.3% CV (RSE 30.0%); bootstrap 45.7 (95% CI 19.7-67.8) |
| `etalcl` IIV CL/F | `log(1 + 0.432^2)` | Table 2: IIV CL/F 43.2% CV (RSE 15.2%); bootstrap 41.6 (95% CI 27.1-53.9) |
| `etaiov_ka_td_<n>` IOV Ka_td | `log(1 + 0.328^2)` (1-10) | Table 2: IOV Ka_td 32.8% CV (RSE 51.1%); bootstrap 39.2 (95% CI 12.0-77.0); shared across occasions |
| `propSd` | `0.234` | Table 2: proportional residual error 23.4% CV (RSE 5.17%); bootstrap 23.2 (95% CI 20.6-25.6) |
| Concentration units | ng/mL | Methods (UPLC-MS/MS; calibration range 0.100-10.0 ng/mL); Table 2 |
| Reference subject | 70 kg adult | Table 2 footnote (CL_70kg / F and V_70kg / F headings) |

# Virtual cohort

The published individual-level data are not available. The cohort below
approximates the Oosten 2016 Table 1 demographics: 52 adults, 63% male /
37% female, body weight drawn from a normal distribution truncated so
that the median is near 75 kg (back-computed from the published BMI
median of 25 kg/m^2 at a typical adult height of 1.73 m, since individual
weights are not tabulated). Every simulated subject receives the median
transdermal patch dose of 50 ug/h.

```{r cohort}
set.seed(20160114) # paper online publication date
n_subj <- 52

cohort <- tibble(
  id        = seq_len(n_subj),
  WT        = pmin(pmax(rnorm(n_subj, mean = 75, sd = 14), 50), 110),
  treatment = factor("50 ug/h matrix patch")
)
```

The transdermal patch is parameterised as a bolus dose into the `depot2`
compartment at each patch application, with first-order release rate
`ka_td = 0.0135 1/h` and lag time 4.73 h. The bolus amount equals the
labelled patch delivery rate multiplied by the wear time:
`50 ug/h * 72 h = 3600 ug` per patch.

```{r events-td}
patch_rate_ug_h  <- 50
patch_interval_h <- 72
patch_dose_ug    <- patch_rate_ug_h * patch_interval_h # 3600 ug per patch

# Single-patch simulation to replicate Figure 1 (stochastic simulation of
# fentanyl concentrations after td 50 ug/h patch application in 52 patients).
obs_times <- sort(unique(c(seq(0, 96, by = 0.5), seq(96, 168, by = 2))))

dose_rows <- cohort |>
  dplyr::mutate(time = 0, amt = patch_dose_ug, cmt = "depot2", evid = 1L, OCC = 1L)

obs_rows <- cohort |>
  tidyr::crossing(time = obs_times) |>
  dplyr::mutate(amt = 0, cmt = NA_character_, evid = 0L, OCC = 1L)

events_td <- dplyr::bind_rows(dose_rows, obs_rows) |>
  dplyr::select(id, time, amt, cmt, evid, WT, OCC, treatment) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(!anyDuplicated(unique(events_td[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate-td}
mod <- rxode2::rxode2(readModelDb("Oosten_2016_fentanyl"))
conc_unit <- mod$units[["concentration"]]
sim_td <- rxode2::rxSolve(
  mod, events = events_td,
  keep = c("WT", "treatment", "OCC")
)
```

# Replicate published figures

## Figure 1 -- stochastic simulation of a 50 ug/h transdermal patch

Oosten 2016 Figure 1 shows the stochastic simulation of fentanyl plasma
concentrations versus time after application of a 50 ug/h transdermal
patch in 52 patients. The simulated 10-50-90 percentile envelope below
should match the figure's reported variability.

```{r figure-1}
fig1 <- sim_td |>
  dplyr::filter(!is.na(Cc), time <= 168) |>
  dplyr::group_by(time, treatment) |>
  dplyr::summarise(
    Q10 = quantile(Cc, 0.10, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q90 = quantile(Cc, 0.90, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(fig1, aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.25) +
  geom_line(linewidth = 0.7) +
  labs(x = "Time after patch application (h)",
       y = paste0("Fentanyl plasma concentration (", conc_unit, ")"),
       title = "Figure 1 -- single 50 ug/h patch, 52 simulated patients",
       caption = "Replicates Oosten 2016 Figure 1.") +
  theme_minimal()
```

## Typical-value Tmax and elimination half-life

The Oosten 2016 Discussion states that for a typical patient receiving a
transdermal patch the predicted Tmax is approximately 20.5 h, the
elimination half-life is 3.91 h, and the absorption half-life is 51.3 h.
The deterministic typical-value prediction below verifies these values
for a 70 kg patient receiving a single 50 ug/h patch (3600 ug bolus into
`depot2`).

```{r typical-td}
mod_typical <- mod |> rxode2::zeroRe()
typical_ev <- rxode2::et(amt = patch_dose_ug, cmt = "depot2") |>
  rxode2::et(amt = 0, evid = 0, time = seq(0, 168, by = 0.25))
typical_ev$WT  <- 70
typical_ev$OCC <- 1L
typical_sim <- rxode2::rxSolve(mod_typical, typical_ev)

# Older rxode2 versions don't always expose the residual-error endpoint Cc as
# an output column after zeroRe(); fall back to central/Vc directly. Vc at the
# reference 70 kg is 280 L (lvc <- fixed(log(280)), e_wt_vc = 1, WT = 70).
cc_vec <- if (!is.null(typical_sim$Cc)) typical_sim$Cc else typical_sim$central / 280
finite_idx <- which(is.finite(cc_vec))
if (length(finite_idx) > 0L) {
  peak_idx  <- finite_idx[which.max(cc_vec[finite_idx])]
  tmax_h    <- typical_sim$time[peak_idx]
  cmax_ngmL <- cc_vec[peak_idx]
} else {
  tmax_h    <- NA_real_
  cmax_ngmL <- NA_real_
}
thalf_el  <- log(2) * 280 / 49.6  # ln(2) * V/CL at 70 kg
thalf_a   <- log(2) / 0.0135       # td absorption half-life

knitr::kable(tibble::tibble(
  Quantity = c("Tmax (h)", "Cmax (ng/mL)", "Elimination half-life (h)",
               "Transdermal absorption half-life (h)"),
  Predicted = c(sprintf("%.1f", tmax_h),
                sprintf("%.3f", cmax_ngmL),
                sprintf("%.2f", thalf_el),
                sprintf("%.1f", thalf_a)),
  `Oosten 2016 reported` = c("~20.5 (Discussion)", "Not directly tabulated",
                             "3.91 (Discussion)", "51.3 (Discussion)")
), caption = "Typical-value verification for a 70 kg adult, single 50 ug/h td patch.")
```

## Figure 4 -- rotation from sc 50 ug/h to a td 50 ug/h patch (12-h scheme)

Oosten 2016 Figure 4 simulates plasma fentanyl during rotation from a sc
infusion of 50 ug/h at steady state to a td patch with delivery rate
50 ug/h using the 12-h scheme: after patch application, sc is continued
at 100% for 6 h, then tapered to 50% for the next 6 h, then stopped.

```{r events-rotation}
# Steady-state sc infusion before rotation (start at -240 h so by t = 0 the
# system is at sc steady state; 240 h is roughly 60 * sc absorption half-life
# of 19.4 h, plenty for equilibration).
ss_h    <- 240
sc_rate <- 50  # ug/h

cohort_rot <- tibble(
  id        = seq_len(n_subj),
  WT        = pmin(pmax(rnorm(n_subj, mean = 75, sd = 14), 50), 110),
  treatment = factor("Rotation 12-h scheme")
)

# Rotation dose schedule, all times in absolute hours from rotation time = 0.
# sc piecewise rates: 50 ug/h for [-ss_h, +6); 25 ug/h for [+6, +12); 0 thereafter.
# td patch (3600 ug bolus to depot2 with OCC = 1) at t = 0.
sc_phase1 <- cohort_rot |>
  dplyr::mutate(time = -ss_h, amt = sc_rate * (ss_h + 6),
                rate = sc_rate, cmt = "depot", evid = 1L, OCC = 0L)
sc_phase2 <- cohort_rot |>
  dplyr::mutate(time = 6, amt = (sc_rate / 2) * 6,
                rate = sc_rate / 2, cmt = "depot", evid = 1L, OCC = 0L)
td_patch <- cohort_rot |>
  dplyr::mutate(time = 0, amt = patch_dose_ug,
                rate = NA_real_, cmt = "depot2", evid = 1L, OCC = 1L)

rot_obs_times <- sort(unique(c(seq(-24, 96, by = 1))))
obs_rot <- cohort_rot |>
  tidyr::crossing(time = rot_obs_times) |>
  dplyr::mutate(amt = 0, rate = NA_real_, cmt = NA_character_,
                evid = 0L, OCC = 1L)

events_rot <- dplyr::bind_rows(sc_phase1, sc_phase2, td_patch, obs_rot) |>
  dplyr::select(id, time, amt, rate, cmt, evid, WT, OCC, treatment) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(!anyDuplicated(unique(events_rot[, c("id", "time", "evid")])))

sim_rot <- rxode2::rxSolve(mod, events = events_rot,
                           keep = c("WT", "treatment", "OCC"))
```

```{r figure-4}
fig4 <- sim_rot |>
  dplyr::filter(!is.na(Cc), time >= -24, time <= 96) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    Q10 = quantile(Cc, 0.10, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q90 = quantile(Cc, 0.90, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(fig4, aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.25) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  labs(x = "Time relative to patch application (h)",
       y = paste0("Fentanyl plasma concentration (", conc_unit, ")"),
       title = "Figure 4 -- sc 50 ug/h -> td 50 ug/h rotation, 12-h scheme",
       caption = "Replicates Oosten 2016 Figure 4. Dashed line = patch application time.") +
  theme_minimal()
```

Oosten 2016 reports for this rotation scheme that 12 h after patch
application the 10th and 90th percentiles are 0.87 and 3.22 ng/mL
respectively, with a median of 1.68 ng/mL. The simulated values below
should be within sampling noise of these.

```{r rotation-checkpoint}
rot_12h <- sim_rot |>
  dplyr::filter(time == 12) |>
  dplyr::summarise(
    Q10 = quantile(Cc, 0.10, na.rm = TRUE),
    median = quantile(Cc, 0.50, na.rm = TRUE),
    Q90 = quantile(Cc, 0.90, na.rm = TRUE)
  )

knitr::kable(tibble::tibble(
  Percentile = c("10th", "Median", "90th"),
  Simulated = sprintf("%.2f", c(rot_12h$Q10, rot_12h$median, rot_12h$Q90)),
  `Oosten 2016 Figure 4 caption / Results` = c("0.87", "1.68", "3.22")
), caption = "Plasma concentration 12 h after patch application (rotation 12-h scheme).")
```

# PKNCA validation

PKNCA single-dose NCA over the 168 h post-patch window for the
stochastic td-only cohort. The treatment grouping comes before `id`
in the formula per project convention. Half-life is computed from the
declining post-Cmax phase, which under the slow-absorption / fast-
elimination regime reflects absorption rather than elimination
(`flip-flop` kinetics, Oosten 2016 Discussion).

```{r pknca}
sim_nca <- sim_td |>
  dplyr::filter(!is.na(Cc), time > 0) |>
  dplyr::select(id, time, Cc, treatment)

dose_df <- events_td |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "ng/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "ug")

intervals <- data.frame(
  start      = 0,
  end        = 168,
  cmax       = TRUE,
  tmax       = TRUE,
  auclast    = TRUE,
  half.life  = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- PKNCA::pk.nca(nca_data)
knitr::kable(summary(nca_res),
  caption = "Single-dose NCA over 168 h post-patch (50 ug/h Durogesic matrix, n = 52).")
```

# Assumptions and deviations

- **Allometric exponents inferred at canonical values.** Oosten 2016
  states "allometrically scaled body weight on CL/F and V/F was found to
  explain some variability and was kept to increase model stability"
  (Results, Fentanyl pharmacokinetics) but does not numerically list the
  exponents. The library model uses the canonical theoretical-allometric
  values 0.75 on CL/F and 1.0 on V/F because (a) the Table 2 headings
  "CL_70kg / F" and "V_70kg / F" imply a `(WT / 70 kg)` normalization
  consistent with the canonical idiom; (b) the paper does not estimate or
  report alternative exponents; and (c) V/F is fixed in the structural
  model, so any custom exponent on V/F would not change the V/F estimate.
  Encoded with `fixed(0.75)` and `fixed(1)` so users see the assumption.
- **V/F fixed.** V/F was fixed to 280 L/70 kg in the paper (Oosten 2016
  Table 2 + Results citation [25]) because the sparse-sampling design did
  not support estimation of all absorption + disposition parameters; a
  sensitivity analysis varying V/F by +/-50% in 10% increments showed
  the model insensitive to the value, with only Tlag_td and Ka_sc varying
  slightly. The library model preserves V/F = 280 L/70 kg as `fixed()`.
- **F_td anchored at 1 at the population level.** Oosten 2016 Table 2
  lists only the IIV on F_td (42.3% CV) without a separately tabulated
  typical value. The library model encodes the typical F_td as
  `fixed(log(1))` and allows the IIV around that anchor; the reference
  subcutaneous route has its bioavailability implicitly anchored at 1 by
  leaving `f(depot)` unset. This is the conventional popPK convention
  for a reference route -- the model is parameterised in apparent
  clearance (CL/F) and apparent volume (V/F), so absolute bioavailability
  is unidentifiable without an absolute reference dose.
- **IOV multiplexed to 10 transdermal occasions.** Oosten 2016 defines an
  occasion as "a transdermal dose followed by at least one observation"
  and the patient cohort had a variable number of patch occasions; the
  library model encodes the per-occasion IOV variance via a fixed set of
  ten occasion-indicator etas (`etaiov_ka_td_1` .. `etaiov_ka_td_10`),
  with the variance estimated for occasion 1 and `fix()`-ed equal across
  occasions 2..10 (NONMEM `$OMEGA BLOCK(1) SAME` translation). The cap
  reflects typical hospital-stay simulations: patches change every 72 h
  so 10 occasions span a 30-day window. Records with `OCC` outside 1..10
  zero every indicator and yield the typical-value Ka_td (no IOV
  applied), which is appropriate for non-transdermal records and for
  simulations beyond ten patch occasions. Users who need more occasions
  can extend the IOV block.
- **Residual error: "additive on the log-scale" -> proportional.** The
  paper's parameterization in NONMEM treats `eps` as additive on the
  log-transformed concentration; this maps cleanly to a proportional
  error model in nlmixr2's linear-concentration space for small-to-
  moderate CV (here 23.4%), and that is the encoding used.
- **Race / ethnicity neutral cohort.** Race is reported in Oosten 2016
  Table 1 (90% Caucasian, 2% other, 8% unknown) but is not a model
  covariate; the simulated cohort is neutral on race.
- **Weight distribution.** Individual body weights were not tabulated in
  Oosten 2016; only the median BMI of 25 kg/m^2 was reported (range
  18-40). The vignette samples body weight from a normal distribution
  centred at 75 kg (back-computed from the BMI median at 1.73 m typical
  height) with SD 14 kg, truncated to [50, 110] kg. Users with their own
  trial cohort should override the cohort generation.
- **Units alignment.** Dosing is parameterised in micrograms (`ug`) and
  concentration in `ng/mL`. Because `1 ug / 1 L = 1 ng/mL`, no explicit
  scaling factor is needed in `model()`; `checkModelConventions()`
  surfaces an `info` flag noting the apparent numerator difference but
  the units are dimensionally consistent.
