---
title: "Daclatasvir + asunaprevir popPK in Japanese HCV (Osawa 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Daclatasvir + asunaprevir popPK in Japanese HCV (Osawa 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Osawa 2018 reports **two independent one-compartment population PK models** --
one per drug -- fitted to overlapping but different cohorts of Japanese patients
with chronic genotype-1 hepatitis C. They are packaged as two model files, per
the library's replicate-the-author's-structure policy, and share this vignette.

```{r load-models}
dcv <- rxode2::rxode(readModelDb("Osawa_2018_daclatasvir"))
asv <- rxode2::rxode(readModelDb("Osawa_2018_asunaprevir"))
```

* Citation: `r dcv$reference`
* Article: <https://doi.org/10.1002/jcph.1274>
* PMC full text: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6174986/>

**Daclatasvir** -- `r dcv$description`

**Asunaprevir** -- `r asv$description`

The daclatasvir and asunaprevir combination ("DUAL") received the first global
approval in Japan in 2014 as the first non-ribavirin, all-oral hepatitis C
therapy, and this analysis is the first model development integrating the
Japanese phase-3 data for both agents.

# Population

Daclatasvir was fitted to **3801 concentrations from 336 subjects** across four
trials (AI444021, AI444022, AI447017, AI447026); asunaprevir to **2626
concentrations from 265 subjects** across two (AI447017, AI447026). All subjects
were Japanese, aged 20 years or older, with chronic genotype-1 HCV (99.4 %
genotype 1b in the daclatasvir cohort). Compensated cirrhosis was permitted and
present in 6.5 % (daclatasvir) and 8.3 % (asunaprevir) of subjects; both cohorts
were about two-thirds female. Daclatasvir was given at 10 or 60 mg once daily,
either with asunaprevir or with peginterferon-alfa/ribavirin; asunaprevir at 200
or 600 mg twice daily as a phase-2 film-coated tablet, or 100 mg twice daily as
the phase-3 soft-gel capsule. Baseline demographics are Osawa 2018 Table 1.

Concentrations below the lower limit of quantification were **excluded**, not
imputed (7 of 3808 records for daclatasvir, 50 of 2676 for asunaprevir).

```{r population}
str(dcv$population, max.level = 1, give.attr = FALSE)
```

# Source trace

Every `ini()` entry carries an in-file comment naming its origin. The table below
collects them, and the chunk that follows **asserts the tabulated values against
the packaged models** so this table cannot silently drift from the code.

```{r source-trace}
trace <- tibble::tribble(
  ~Model,         ~Parameter,                  ~Value,  ~`Source location`,
  "daclatasvir",  "lcl (CL/F, L/h)",            5.29,   "Table 2, daclatasvir theta1",
  "daclatasvir",  "lvc (V/F, L)",              64.2,    "Table 2, daclatasvir theta2",
  "daclatasvir",  "lka (Ka, 1/h)",              0.865,  "Table 2, daclatasvir theta3",
  "daclatasvir",  "e_sexf_cl",                 -0.110,  "Table 2, daclatasvir theta6",
  "daclatasvir",  "e_conmed_pegifn_rbv_cl",    -0.122,  "Table 2, daclatasvir theta8",
  "daclatasvir",  "e_crcl_base_cl",             0.235,  "Table 2, daclatasvir theta10",
  "daclatasvir",  "e_wt_base_vc",               0.605,  "Table 2, daclatasvir theta14",
  "daclatasvir",  "expSd (log-scale SD)",       0.375,  "Table 2, daclatasvir theta4",
  "asunaprevir",  "lcl (CL/F, L/h)",           52.1,    "Table 2, asunaprevir theta1",
  "asunaprevir",  "lvc (V/F, L)",              75.1,    "Table 2, asunaprevir theta2",
  "asunaprevir",  "lka (Ka, 1/h)",              0.228,  "Table 2, asunaprevir theta3",
  "asunaprevir",  "e_ast_base_cl",             -0.598,  "Table 2, asunaprevir theta7",
  "asunaprevir",  "e_ast_cl",                  -0.382,  "Table 2, asunaprevir theta8",
  "asunaprevir",  "e_form_asv_softgel_cl_vc",  -0.314,  "Table 2, asunaprevir theta10 ('F ~ FORM')",
  "asunaprevir",  "e_dis_cirrhosis_cl",        -0.428,  "Table 2, asunaprevir theta11",
  "asunaprevir",  "expSd (log-scale SD)",       0.68,   "Table 2, asunaprevir theta4"
)

# Gate: each tabulated value must equal the packaged model's estimate. `l`-
# prefixed structural parameters are stored on the log scale, so exponentiate.
ini_value <- function(ui, nm) {
  row <- ui$iniDf[!is.na(ui$iniDf$name) & ui$iniDf$name == nm, ]
  stopifnot(nrow(row) == 1L)
  row$est
}
uis <- list(daclatasvir = dcv, asunaprevir = asv)
trace_check <- trace |>
  mutate(
    param = sub(" .*$", "", Parameter),
    packaged = unname(vapply(
      seq_len(n()),
      function(i) {
        p <- param[i]
        v <- ini_value(uis[[Model[i]]], p)
        # Structural PK parameters are stored log-transformed.
        if (p %in% c("lcl", "lvc", "lka")) exp(v) else v
      },
      numeric(1)
    )),
    rel_diff = abs(packaged - Value) / pmax(abs(Value), 1e-12)
  )
stopifnot(
  nrow(trace_check) == 16L,
  !anyNA(trace_check$rel_diff),
  max(trace_check$rel_diff) < 1e-10
)

knitr::kable(
  trace,
  caption = "Source trace. Every value is asserted against the packaged model above."
)
```

Random effects (Osawa 2018 Table 2, "Random effects" block, read as
*variance (standard deviation)* on diagonals and *covariance (correlation)*
off-diagonal, per footnote b):

| Model | Term | Variance | Notes |
|---|---|---|---|
| daclatasvir | `etalcl` | 0.155 | SD 0.394 |
| daclatasvir | `etalvc` | 0.145 | SD 0.381 |
| daclatasvir | `etalcl`:`etalvc` | 0.141 | correlation 0.941 |
| daclatasvir | `etalka` | 0.756 | SD 0.869 |
| daclatasvir | `etaexpSd` | 0.107 | SD 0.327; IIV on the residual magnitude |
| asunaprevir | `etalcl` | 0.172 | SD 0.415 |
| asunaprevir | `etalvc` | 0.872 | SD 0.934 |
| asunaprevir | `etaexpSd` | 0.0672 | SD 0.259; IIV on the residual magnitude |

Two structural features of that block are worth spelling out, because both are
easy to mis-transcribe:

1. **The residual error carries its own eta.** Table 2 prints `sigma` twice --
   once as `theta4` under "Residual error" and once as `omega4,4` under "Random
   effects" -- with different standard errors (RSE 2.72 % vs 18.0 % for
   daclatasvir). These are two separate estimated parameters, not one value
   printed twice. The Methods residual model is log-transform-both-sides,
   `ln(y) = ln(yhat) + theta_ADD * eps`, so the packaged models use nlmixr2's
   `lnorm()` structure with a per-subject SD `expSd * exp(etaexpSd)` -- the
   NONMEM `W = THETA(4) * EXP(ETA(4))` construct.
2. **Asunaprevir Ka carries no eta at all.** "The random effect of Ka was fixed
   as zero because the sampling points around the peak concentrations were not
   enough for all subjects." A zero-variance eta would make OMEGA singular, so
   the eta is omitted rather than encoded as `fixed(0)`.

## Covariate equations

Transcribed verbatim from the Results section (the PDF renders these as images,
so they are absent from text-extraction of the article; they were recovered with
`pdftotext -layout`). Daclatasvir:

$$
\mathrm{CL/F} = \mathrm{CL/F_{ref}} \cdot
  \left(\frac{\mathrm{BCRCL}}{86.48}\right)^{0.235} \cdot
  \exp\left(-0.110\cdot\mathrm{SEX} - 0.122\cdot\mathrm{TX}\right),
\qquad
\mathrm{V/F} = \mathrm{V/F_{ref}} \cdot
  \left(\frac{\mathrm{BBWT}}{56}\right)^{0.605}
$$

with reference male, DUAL regimen, 86.48 mL/min, 56 kg. Asunaprevir:

$$
\mathrm{CL/F} = \mathrm{CL/F_{ref}} \cdot
  \left(\frac{\mathrm{BAST}}{52}\right)^{-0.598}
  \left(\frac{\mathrm{AST}}{52}\right)^{-0.382} \cdot
  \exp\left(-0.314\cdot\mathrm{FORM} - 0.428\cdot\mathrm{CIRRHOSIS}\right),
\qquad
\mathrm{V/F} = \mathrm{V/F_{ref}} \cdot \exp\left(-0.314\cdot\mathrm{FORM}\right)
$$

with reference phase-2 tablet, no cirrhosis, and **52 U/L for both AST terms** --
the printed equation normalises the time-varying AST by `BASTref`, not by a
separate reference.

The single coefficient `-0.314` appears in both asunaprevir equations because
Table 2 tabulates it once, labelled `F ~ FORM`: bioavailability is not separately
identifiable from oral data, so an F effect enters the apparent quantities CL/F
and V/F identically. Scaling both leaves CL/V untouched and raises concentrations
by `1/exp(-0.314)`, which is checked below.

# Structural checks

All checks in this section are **deterministic** -- run with random effects
zeroed -- so they use tight tolerances. They would break on a mis-transcribed
parameter, reference constant, or equation form.

```{r helpers}
# Typical-value (zeroRe) solve at a stated covariate set.
solve_typical <- function(ui, covariates, dose, ii, n_doses, obs_times) {
  ev <- rxode2::et(amt = dose, ii = ii, addl = n_doses - 1L, cmt = "depot")
  ev <- rxode2::et(ev, obs_times, cmt = "central")
  d <- as.data.frame(ev)
  for (nm in names(covariates)) d[[nm]] <- covariates[[nm]]
  suppressWarnings(
    as.data.frame(rxode2::rxSolve(rxode2::zeroRe(ui), d, returnType = "data.frame"))
  )
}
trapz <- function(t, y) sum(diff(t) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)

dcv_ref <- list(CRCL_BASE = 86.48, WT_BASE = 56, SEXF = 0, CONMED_PEGIFN_RBV = 0)
asv_ref <- list(AST_BASE = 52, AST = 52, DIS_CIRRHOSIS = 0, FORM_ASV_SOFTGEL = 0)
```

## 1. Typical values reproduce Table 2

```{r check-typical}
dcv_tv <- solve_typical(dcv, dcv_ref, dose = 60, ii = 24, n_doses = 20,
                        obs_times = seq(0, 480, by = 0.1))
asv_tv <- solve_typical(asv, asv_ref, dose = 200, ii = 12, n_doses = 40,
                        obs_times = seq(0, 480, by = 0.1))

stopifnot(
  isTRUE(all.equal(unique(dcv_tv$cl), 5.29,  tolerance = 1e-10)),
  isTRUE(all.equal(unique(dcv_tv$vc), 64.2,  tolerance = 1e-10)),
  isTRUE(all.equal(unique(dcv_tv$ka), 0.865, tolerance = 1e-10)),
  isTRUE(all.equal(unique(asv_tv$cl), 52.1,  tolerance = 1e-10)),
  isTRUE(all.equal(unique(asv_tv$vc), 75.1,  tolerance = 1e-10)),
  isTRUE(all.equal(unique(asv_tv$ka), 0.228, tolerance = 1e-10))
)
```

At the reference covariates both models return the Table 2 estimates exactly.

## 2. Steady-state mass balance: `CL/F * AUCtau == Dose`

For a linear model with the whole dose entering the depot, the steady-state area
over one dosing interval must satisfy `AUCtau = Dose / (CL/F)` exactly. The only
error here is trapezoidal, so a tight bound is correct (the two sides use the
same parameters; this is a solve against its own closed form, not a comparison
across a physical mechanism).

```{r check-mass-balance}
ss_auc_err <- function(sim, t_start, t_end, dose) {
  w <- sim[sim$time >= t_start & sim$time <= t_end, ]
  abs(unique(w$cl) * trapz(w$time, w$Cc) - dose) / dose
}
err_dcv <- ss_auc_err(dcv_tv, 456, 480, 60)   # last 24 h interval, QD
err_asv <- ss_auc_err(asv_tv, 468, 480, 200)  # last 12 h interval, BID

# Realised ~2e-6 (daclatasvir) and ~1e-5 (asunaprevir) on a 0.1 h grid.
stopifnot(err_dcv < 1e-3, err_asv < 1e-3)
c(daclatasvir = err_dcv, asunaprevir = err_asv)
```

## 3. Closed-form one-compartment oral solution

An independent check that the ODE system is the intended structure: after a
single dose, `Cc(t) = (D * ka) / (V * (ka - kel)) * (exp(-kel*t) - exp(-ka*t))`.

```{r check-closed-form}
closed_form <- function(t, dose, ka, cl, vc) {
  kel <- cl / vc
  dose * ka / (vc * (ka - kel)) * (exp(-kel * t) - exp(-ka * t))
}
sd_times <- seq(0, 120, by = 0.1)
dcv_sd <- solve_typical(dcv, dcv_ref, dose = 60, ii = 24, n_doses = 1, obs_times = sd_times)
asv_sd <- solve_typical(asv, asv_ref, dose = 200, ii = 12, n_doses = 1, obs_times = sd_times)

max_rel <- function(sim, dose) {
  cf <- closed_form(sim$time, dose, unique(sim$ka), unique(sim$cl), unique(sim$vc))
  keep <- cf > 1e-8
  max(abs(sim$Cc[keep] - cf[keep]) / cf[keep])
}
cf_dcv <- max_rel(dcv_sd, 60)
cf_asv <- max_rel(asv_sd, 200)
stopifnot(cf_dcv < 1e-4, cf_asv < 1e-4)
c(daclatasvir = cf_dcv, asunaprevir = cf_asv)
```

## 4. Asunaprevir is flip-flop

`Ka = 0.228 1/h` sits well below `kel = CL/F / (V/F) = 0.694 1/h`, so the
terminal slope of the asunaprevir profile reflects **absorption**, not
elimination. Daclatasvir is the ordinary case.

```{r check-flipflop}
rates <- tibble::tibble(
  Drug = c("daclatasvir", "asunaprevir"),
  ka = c(unique(dcv_tv$ka), unique(asv_tv$ka)),
  kel = c(unique(dcv_tv$cl) / unique(dcv_tv$vc), unique(asv_tv$cl) / unique(asv_tv$vc))
) |>
  mutate(
    `t1/2 from ka (h)` = log(2) / ka,
    `t1/2 from kel (h)` = log(2) / kel,
    `Flip-flop` = ka < kel,
    `Terminal t1/2 (h)` = log(2) / pmin(ka, kel)
  )
stopifnot(
  rates$`Flip-flop` == c(FALSE, TRUE),
  # Terminal half-life is set by the slower of the two rate constants.
  isTRUE(all.equal(rates$`Terminal t1/2 (h)`, c(log(2) / rates$kel[1], log(2) / rates$ka[2])))
)
knitr::kable(rates, digits = 3, caption = "Absorption vs elimination rate constants.")
```

## 5. Covariate effects reproduce the paper's forest plot

Osawa 2018 states each covariate's effect at the 5th and 95th percentiles of the
covariate (Figure 3 and the "Effect of Covariates" text). Reproducing those
numbers is the single most informative check on the covariate equations: a wrong
reference constant, a sign flip, or a power-versus-exponential mix-up all break
it. Because the fold-changes are deterministic typical-value ratios, the
tolerances are tight.

```{r check-covariates}
fold <- function(ui, ref, changes, param, ref_value) {
  covs <- utils::modifyList(ref, changes)
  s <- solve_typical(ui, covs, dose = 1, ii = 24, n_doses = 1, obs_times = c(0, 1))
  unique(s[[param]]) / ref_value
}

cov_checks <- tibble::tribble(
  ~Drug, ~Covariate, ~`Paper claim`, ~Published, ~Tol, ~Achieved,
  "daclatasvir", "BBWT 42.8 kg (5th pct) on V/F", "~15% lower",  0.850, 0.010,
    fold(dcv, dcv_ref, list(WT_BASE = 42.8), "vc", 64.2),
  "daclatasvir", "BBWT 78 kg (95th pct) on V/F",  "~22% higher", 1.220, 0.010,
    fold(dcv, dcv_ref, list(WT_BASE = 78), "vc", 64.2),
  "daclatasvir", "BCRCL 51.36 (5th pct) on CL/F", "~10% lower",  0.900, 0.030,
    fold(dcv, dcv_ref, list(CRCL_BASE = 51.36), "cl", 5.29),
  "daclatasvir", "BCRCL 144.42 (95th pct) on CL/F", "~10% higher", 1.100, 0.035,
    fold(dcv, dcv_ref, list(CRCL_BASE = 144.42), "cl", 5.29),
  "daclatasvir", "Female on CL/F",               "~10% lower",  exp(-0.110), 0.001,
    fold(dcv, dcv_ref, list(SEXF = 1), "cl", 5.29),
  "daclatasvir", "pegIFN/RBV backbone on CL/F",  "~10% lower",  exp(-0.122), 0.001,
    fold(dcv, dcv_ref, list(CONMED_PEGIFN_RBV = 1), "cl", 5.29),
  "asunaprevir", "BAST 22 U/L (5th pct) on CL/F", "1.67-fold higher", 1.67, 0.010,
    fold(asv, asv_ref, list(AST_BASE = 22), "cl", 52.1),
  "asunaprevir", "BAST 123.6 U/L (95th pct) on CL/F", "0.60-fold", 0.60, 0.010,
    fold(asv, asv_ref, list(AST_BASE = 123.6), "cl", 52.1),
  "asunaprevir", "AST 16 U/L (5th pct) on CL/F", "1.57-fold higher", 1.57, 0.010,
    fold(asv, asv_ref, list(AST = 16), "cl", 52.1),
  "asunaprevir", "AST 115 U/L (95th pct) on CL/F", "0.74-fold", 0.74, 0.010,
    fold(asv, asv_ref, list(AST = 115), "cl", 52.1),
  "asunaprevir", "Cirrhosis on CL/F",            "0.65-fold", 0.65, 0.010,
    fold(asv, asv_ref, list(DIS_CIRRHOSIS = 1), "cl", 52.1)
) |>
  mutate(
    Difference = Achieved - Published,
    Pass = abs(Difference) <= Tol
  )

stopifnot(is.logical(cov_checks$Pass), !anyNA(cov_checks$Pass), all(cov_checks$Pass))

cov_checks |>
  select(Drug, Covariate, `Paper claim`, Published, Achieved, Difference, Pass) |>
  knitr::kable(digits = 4, caption = "Covariate effects vs the values printed in Osawa 2018.")
```

Every effect reproduces. The two baseline-creatinine-clearance rows carry a wider
tolerance because the paper reports them only as "approximately 10 %"; the
encoded power function returns 11.5 % and 12.8 %, which round to the paper's
one-significant-figure claim. Every other row matches the published figure to
within 0.01.

## 6. Asunaprevir relative bioavailability is 1.37-fold

```{r check-relative-f}
asv_softgel <- solve_typical(
  asv, utils::modifyList(asv_ref, list(FORM_ASV_SOFTGEL = 1)),
  dose = 100, ii = 12, n_doses = 1, obs_times = c(0, 1)
)
rel_f <- 52.1 / unique(asv_softgel$cl)
kel_tablet <- unique(asv_tv$cl) / unique(asv_tv$vc)
kel_softgel <- unique(asv_softgel$cl) / unique(asv_softgel$vc)

stopifnot(
  # Paper: "bioavailability of the phase 3 soft-gel capsule ... 1.37-fold higher".
  abs(rel_f - 1.37) < 0.005,
  # The formulation factor must cancel out of CL/V, leaving the rate constant
  # -- and hence the profile SHAPE -- formulation-independent.
  isTRUE(all.equal(kel_tablet, kel_softgel, tolerance = 1e-10))
)
c(`relative F (softgel/tablet)` = rel_f, `kel tablet` = kel_tablet, `kel softgel` = kel_softgel)
```

# Virtual cohort

Individual data are not public ("The individual deidentified participant data and
related documents will not be shared"), so the simulations below use virtual
cohorts whose covariate distributions match the medians and 5th/95th percentiles
Osawa 2018 reports. Log-normal distributions calibrated on the published median
and 95th percentile reproduce the published 5th percentile closely -- 40.2 vs
42.8 kg for weight, 51.8 vs 51.36 mL/min for creatinine clearance, 21.9 vs 22 U/L
for baseline AST -- which is itself a mild consistency check on the cohort.

```{r cohort}
# set.seed() seeds R's RNG, not rxode2's; rxode2 streams are partitioned per
# solver thread, so the cohort differs between a 2-core CI runner and a
# workstation. Every assertion downstream is written to hold for ANY cohort.
set.seed(20181101)
rxode2::rxSetSeed(20181101)

N_ARM <- 150L  # per arm; cohorts are capped at 200/arm

# Log-normal calibrated so median = m and 95th percentile = p95, truncated to
# the published observed range.
rlnorm_pct <- function(n, m, p95, lo, hi) {
  pmin(pmax(stats::rlnorm(n, log(m), log(p95 / m) / stats::qnorm(0.95)), lo), hi)
}

make_arm <- function(n, drug, arm, dose, ii, n_doses, obs_grid, id_offset) {
  subj <- tibble::tibble(id = id_offset + seq_len(n), arm = arm, drug = drug)
  if (drug == "daclatasvir") {
    subj <- subj |>
      mutate(
        WT_BASE = rlnorm_pct(n, 56, 78, 36, 93),           # Table 1 median 56, p95 78
        CRCL_BASE = rlnorm_pct(n, 86.48, 144.42, 39.56, 185.96),
        SEXF = stats::rbinom(n, 1, 0.637),                  # Table 1: 63.7% female
        CONMED_PEGIFN_RBV = stats::rbinom(n, 1, 0.211)      # Table 1: 21.1% on pegIFN/RBV
      )
  } else {
    subj <- subj |>
      mutate(
        AST_BASE = rlnorm_pct(n, 52, 123.6, 13, 595),       # Table 1 median 52, p95 123.6
        DIS_CIRRHOSIS = stats::rbinom(n, 1, 0.083),         # Table 1: 8.3% cirrhotic
        FORM_ASV_SOFTGEL = as.integer(arm == "100 mg BID soft-gel"),
        # Assumption: time-varying AST held at each subject's baseline (the
        # paper publishes no AST trajectory). See Assumptions below.
        AST = AST_BASE
      )
  }
  t_last <- ii * (n_doses - 1L)
  doses <- subj |>
    tidyr::crossing(time = ii * seq.int(0, n_doses - 1L)) |>
    mutate(amt = dose, evid = 1L, cmt = "depot")
  obs <- subj |>
    tidyr::crossing(time = t_last + obs_grid) |>
    # cmt is the ODE STATE, never the observable name `Cc`.
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}

# Observation grid: log-spaced over the first hour, then regular. A UNIFORM
# grid is not good enough here, because both models draw individuals with very
# fast early kinetics -- daclatasvir's Ka carries omega = 0.756 (SD 0.869), so
# some subjects absorb at ~10 /h, and asunaprevir's V/F carries omega = 0.872
# (SD 0.934), so some subjects have kel = CL/V near 10 /h and a sub-0.1 h
# post-dose transient. On a uniform 0.25 h grid the trapezoidal AUC of those
# subjects is off by up to 1%, which is numerical error rather than a model
# property but is large enough to blunt the mass-balance gate below. Resolving
# the first hour costs almost nothing and drops the error ~30-fold.
obs_grid <- function(tau) {
  sort(unique(c(0, exp(seq(log(0.005), log(1), length.out = 40)), seq(1, tau, by = 0.25))))
}

# 14 days of dosing, then dense observation across the final dosing interval --
# matching Osawa 2018 Figure 2, "the pcVPC plot after 14 days at steady state".
dcv_events <- bind_rows(
  make_arm(N_ARM, "daclatasvir", "60 mg QD", 60, 24, 14L, obs_grid(24), 0L),
  make_arm(N_ARM, "daclatasvir", "10 mg QD", 10, 24, 14L, obs_grid(24), 1000L)
)
asv_events <- bind_rows(
  make_arm(N_ARM, "asunaprevir", "100 mg BID soft-gel", 100, 12, 28L, obs_grid(12), 2000L),
  make_arm(N_ARM, "asunaprevir", "200 mg BID tablet", 200, 12, 28L, obs_grid(12), 3000L)
)

# Disjoint ids across arms: duplicate ids silently merge into one subject that
# receives the summed dose.
stopifnot(
  !anyDuplicated(unique(dcv_events[, c("id", "time", "evid")])),
  !anyDuplicated(unique(asv_events[, c("id", "time", "evid")])),
  length(intersect(dcv_events$id, asv_events$id)) == 0L
)
```

# Simulation

```{r simulate}
dcv_sim <- as.data.frame(rxode2::rxSolve(dcv, dcv_events, keep = c("arm"))) |>
  mutate(tad = time - 24 * 13)
asv_sim <- as.data.frame(rxode2::rxSolve(asv, asv_events, keep = c("arm"))) |>
  mutate(tad = time - 12 * 27)

# `Cc` is the individual prediction; `sim` additionally carries the lnorm
# residual, which is what an observed concentration would look like.
stopifnot(
  all(c("Cc", "sim", "expSdInd") %in% names(dcv_sim)),
  all(dcv_sim$Cc >= 0), all(asv_sim$Cc >= 0)
)
```

The per-subject residual SD varies, because the residual magnitude carries its
own eta (`omega4,4`). Its population median must return the tabulated `theta4`:

```{r check-residual-iiv}
resid_sd <- bind_rows(
  dcv_sim |> distinct(id, expSdInd) |> mutate(Drug = "daclatasvir", theta4 = 0.375),
  asv_sim |> distinct(id, expSdInd) |> mutate(Drug = "asunaprevir", theta4 = 0.68)
) |>
  group_by(Drug, theta4) |>
  summarise(n = n(), median_sd = stats::median(expSdInd), .groups = "drop") |>
  mutate(ratio = median_sd / theta4)

# A cohort median, so the bound admits sampling noise rather than asserting
# equality. With n = 300 per drug and log-SD 0.327 / 0.259 the median of
# exp(eta) sits within a few percent of 1; 15% leaves ample headroom while
# still breaking if the eta were applied on the wrong scale (e.g. as a
# variance, or additively).
stopifnot(all(abs(resid_sd$ratio - 1) < 0.15))
knitr::kable(resid_sd, digits = 4,
             caption = "Median per-subject residual SD vs the tabulated theta4.")
```

# Replicate published figures

```{r figure-2, fig.cap = "Steady-state concentration-time profiles after 14 days, replicating the layout of Osawa 2018 Figure 2 (prediction-corrected VPC). Ribbons span the 5th-95th percentiles of simulated observations; the line is the median."}
vpc_summary <- function(sim) {
  sim |>
    group_by(arm, tad) |>
    summarise(
      Q05 = stats::quantile(sim, 0.05, na.rm = TRUE),
      Q50 = stats::quantile(sim, 0.50, na.rm = TRUE),
      Q95 = stats::quantile(sim, 0.95, na.rm = TRUE),
      .groups = "drop"
    )
}
bind_rows(
  vpc_summary(dcv_sim) |> mutate(Drug = "Daclatasvir"),
  vpc_summary(asv_sim) |> mutate(Drug = "Asunaprevir")
) |>
  ggplot(aes(tad, Q50, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~Drug, scales = "free") +
  scale_y_log10() +
  labs(
    x = "Time after dose on day 14 (h)", y = "Concentration (ug/mL)",
    colour = NULL, fill = NULL,
    caption = "Replicates the steady-state layout of Figure 2 of Osawa 2018."
  ) +
  theme(legend.position = "bottom")
```

```{r figure-3, fig.cap = "Covariate effects on the typical PK parameter, replicating Osawa 2018 Figure 3. The 80-125% reference band is shaded."}
forest <- cov_checks |>
  mutate(
    Parameter = ifelse(grepl("V/F", Covariate), "V/F", "CL/F"),
    Label = sub(" on (CL|V)/F$", "", Covariate)
  )
ggplot(forest, aes(Achieved, Label, colour = Drug)) +
  annotate("rect", xmin = 0.8, xmax = 1.25, ymin = -Inf, ymax = Inf,
           alpha = 0.15, fill = "grey40") +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_point(size = 2.5) +
  facet_grid(Drug ~ ., scales = "free_y", space = "free_y") +
  labs(
    x = "Fold-change vs the typical subject", y = NULL, colour = NULL,
    caption = "Replicates Figure 3 of Osawa 2018. Shaded band is the 80-125% boundary."
  ) +
  theme(legend.position = "none")
```

The figure reproduces the paper's central qualitative finding: **every
daclatasvir covariate effect lies within or overlaps the 80-125 % band**, while
**every asunaprevir effect escapes it**. That contrast is the paper's headline
conclusion, so it is asserted rather than left to the eye:

```{r check-forest-conclusion}
inside_band <- function(x) x >= 0.8 & x <= 1.25
concl <- forest |>
  group_by(Drug) |>
  summarise(n_outside = sum(!inside_band(Achieved)), n = n(), .groups = "drop")

stopifnot(
  # Osawa 2018 Abstract: daclatasvir covariate effects "were within or
  # overlapped the 80% to 125% boundaries"; asunaprevir's "exceeded" them.
  concl$n_outside[concl$Drug == "daclatasvir"] == 0L,
  concl$n_outside[concl$Drug == "asunaprevir"] == concl$n[concl$Drug == "asunaprevir"]
)
knitr::kable(concl, caption = "Covariate effects outside the 80-125% band, by drug.")
```

# PKNCA validation

NCA is run on the individual predictions `Cc` over the final steady-state dosing
interval, with time re-zeroed to the last dose so the interval is `[0, tau]`.

```{r pknca}
run_nca <- function(sim, events, tau, t_last) {
  conc <- sim |>
    filter(!is.na(Cc), tad >= 0, tad <= tau) |>
    transmute(id, arm, time = tad, Cc)
  # Guarantee a time-zero anchor per (id, arm). At steady state the pre-dose
  # trough is a real, non-zero concentration, so carry the simulated value
  # rather than inserting a zero.
  conc <- conc |>
    distinct(id, arm, time, .keep_all = TRUE) |>
    arrange(id, arm, time)
  stopifnot(nrow(conc) > 0L, all(conc |> count(id) |> pull(n) > 1L))

  dose_df <- events |>
    filter(evid == 1L, time == t_last) |>
    transmute(id, arm, time = 0, amt)

  conc_obj <- PKNCA::PKNCAconc(as.data.frame(conc), Cc ~ time | arm + id)
  dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | arm + id)
  intervals <- data.frame(
    start = 0, end = tau,
    cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE
  )
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}

dcv_nca <- run_nca(dcv_sim, dcv_events, tau = 24, t_last = 24 * 13)
asv_nca <- run_nca(asv_sim, asv_events, tau = 12, t_last = 12 * 27)
```

```{r nca-summary}
nca_wide <- function(res, drug) {
  as.data.frame(res$result) |>
    filter(PPTESTCD %in% c("cmax", "tmax", "cmin", "auclast")) |>
    select(arm, id, PPTESTCD, PPORRES) |>
    tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
    mutate(Drug = drug)
}
nca_all <- bind_rows(nca_wide(dcv_nca, "daclatasvir"), nca_wide(asv_nca, "asunaprevir"))

nca_all |>
  group_by(Drug, arm) |>
  summarise(
    n = n(),
    `Cmax (ug/mL)` = stats::median(cmax),
    `Tmax (h)` = stats::median(tmax),
    `Ctrough (ug/mL)` = stats::median(cmin),
    `AUCtau (ug*h/mL)` = stats::median(auclast),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 3, caption = "Simulated steady-state NCA (median across subjects).")
```

## PKNCA AUCtau recovers Dose / (CL/F) per subject

This ties the NCA output back to the closed form, per subject. It is the
strictest available gate on the whole chain -- model, event table, simulation and
NCA setup -- and it is exact up to trapezoidal error, so the bound is tight.

```{r check-nca-recovery}
cl_by_id <- bind_rows(
  dcv_sim |> distinct(id, arm, cl),
  asv_sim |> distinct(id, arm, cl)
)
dose_by_arm <- tibble::tibble(
  arm = c("60 mg QD", "10 mg QD", "100 mg BID soft-gel", "200 mg BID tablet"),
  amt = c(60, 10, 100, 200)
)

recovery <- nca_all |>
  left_join(cl_by_id, by = c("id", "arm")) |>
  left_join(dose_by_arm, by = "arm") |>
  mutate(rel_err = abs(cl * auclast - amt) / amt)

stopifnot(
  nrow(recovery) == 4L * N_ARM,
  !anyNA(recovery$rel_err),
  # The discrepancy here is pure numerical (AUC quadrature) error: both sides
  # use the SAME drawn parameters, so this is a solve checked against its own
  # closed form and a tight bound is correct. Realised max across three seeds
  # on the log-spaced grid: 2.0e-4 / 2.2e-4 / 2.1e-4 (daclatasvir) and
  # 3.6e-4 / 3.6e-4 / 1.2e-3 (asunaprevir, the tail driven by the most extreme
  # V/F draw). 5e-3 leaves ~4x headroom over the worst observed draw while
  # still breaking on any structural error -- a wrong dose, volume, unit or
  # covariate reference moves this by tens of percent, not tenths of one.
  max(recovery$rel_err) < 0.005
)
recovery |>
  group_by(Drug, arm) |>
  summarise(`max |CL*AUCtau - Dose| / Dose` = max(rel_err), .groups = "drop") |>
  knitr::kable(digits = 6, caption = "Per-subject steady-state mass-balance recovery.")
```

## Comparison against published NCA

Osawa 2018 does not publish an NCA table of its own. It does quote observed
asunaprevir exposures for the two contributing trials, carried over from
reference 17: **2950 ng*h/mL in AI447017** (the phase-2 study, 200 mg tablet) and
**2155 ng*h/mL in AI447026** (the phase-3 study, 100 mg soft-gel capsule).

```{r nca-compare}
asv_sim_nca <- nca_all |>
  filter(Drug == "asunaprevir") |>
  group_by(arm) |>
  summarise(cmax = stats::median(cmax), tmax = stats::median(tmax),
            auclast = stats::median(auclast) * 1000, .groups = "drop")  # ug*h/mL -> ng*h/mL

published <- tibble::tribble(
  ~arm,                   ~auclast,
  "200 mg BID tablet",     2950,
  "100 mg BID soft-gel",   2155
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = asv_sim_nca |> select(arm, auclast),
  reference = published,
  by = "arm",
  units = c(auclast = "ng*h/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Simulated steady-state AUCtau vs the observed exposures Osawa 2018 quotes from reference 17. * differs by >20%.")
```

**Both rows are starred, and that is expected rather than a transcription
error.** The model over-predicts by about 30 % (tablet) and 22 % (soft-gel).
Three reasons, none of which is fixable by tuning:

1. **The reference values are not this paper's data.** They are observed NCA
   means quoted from reference 17 for two individual trials, whereas the model
   was fitted to the pooled dataset.
2. **The model's typical subject is not those trials' mean subject.** The
   packaged typical values are defined at baseline AST 52 U/L, time-varying AST
   52 U/L and no cirrhosis. Asunaprevir CL/F has a combined AST exponent of
   -0.980, so a cohort whose mean AST sits above 52 U/L has systematically higher
   clearance and lower exposure than the reference subject.
3. **A median of a right-skewed simulated distribution is being compared with a
   published arithmetic mean.** With `omega(V/F) = 0.872` the asunaprevir
   exposure distribution is very wide.

The internally consistent check the paper *does* let us make is the ratio: the
two quoted exposures differ by a factor of `r round(2950 / 2155, 3)`, and the
fitted formulation effect is `r round(1 / exp(-0.314), 3)`-fold. Those agree to
three decimal places, although at different doses (200 mg vs 100 mg), so the
agreement is suggestive rather than a controlled relative-bioavailability
comparison -- which is how the paper itself treats it ("the trend was consistent
with the result of the relative bioavailability study").

```{r check-published-ratio}
observed_ratio <- 2950 / 2155
fitted_rel_f <- 1 / exp(-0.314)
# Both are fixed published numbers, so this is a transcription check on theta10,
# not a simulation-derived assertion.
stopifnot(abs(observed_ratio - fitted_rel_f) < 0.01)
c(`observed AUC ratio` = observed_ratio, `fitted relative F` = fitted_rel_f)
```

# Assumptions and deviations

**Assumptions made because the paper does not specify.**

- **Time-varying AST is held at each subject's baseline** in the asunaprevir
  virtual cohort. Osawa 2018 publishes no AST trajectory, only the baseline
  distribution and the 5th/95th percentiles of AST at the last sampling time.
  A subject whose AST never moves carries the *sum* of the two exponents,
  -0.980, on `(AST/52)`. This affects the simulated spread of CL/F but not the
  typical value, since both terms equal 1 at the reference. Structural check 5
  exercises each AST term separately, so the covariate equation is validated
  independently of this cohort choice.
- **Covariate distributions** are log-normal (weight, creatinine clearance,
  baseline AST) or Bernoulli (sex, regimen, cirrhosis, formulation), calibrated
  on the published medians, 5th/95th percentiles and category frequencies
  (Table 1 and the "Effect of Covariates" text) and truncated to the published
  observed range. Correlations among covariates are **not** reproduced, although
  the paper states that age, baseline weight and baseline creatinine clearance
  were moderately-to-highly correlated, as were AST and ALT. Only one member of
  each correlated set entered each final model, so the omission does not bias the
  typical-value predictions.
- **Dosing is 14 days of regular administration** with no missed doses, matching
  the "after 14 days at steady state" framing of Figure 2.

**Encoding decisions worth flagging to a reviewer.**

- **The formulation effect is applied to CL/F and V/F, not through `f(depot)`.**
  Table 2 labels `theta10` as an effect on `F`, but the paper's printed equations
  apply that single coefficient to CL/F and to V/F. The two readings are
  mathematically identical for oral-only data; the packaged model transcribes the
  equations as printed so a reviewer can check it line by line, and structural
  check 6 confirms it delivers exactly the 1.37-fold relative bioavailability the
  paper reports while leaving the elimination rate constant untouched.
- **Inter-individual variability on the residual-error magnitude is encoded.**
  Table 2's `omega4,4` row sits in the "Random effects" block under the symbol
  `sigma`, alongside `theta4` in the "Residual error" block. This is the NONMEM
  `W = THETA(4) * EXP(ETA(4))` construct, and both models encode it as
  `expSd * exp(etaexpSd)` passed to `lnorm()`. An extraction that dropped it
  would understate the spread of simulated observations.
- **Asunaprevir `Ka` has no eta.** The authors fixed its random effect to zero;
  a zero-variance eta would make OMEGA singular, so it is omitted from `ini()`
  rather than written as `fixed(0)`.

**Apparent errors in the source, and how they were resolved.**

- **Figure 3 caption prints the baseline creatinine clearance reference as
  "88.48 mL/min".** The equation text, the Table 1 median (86.5) and the printed
  5th/95th-percentile fold-changes all agree on **86.48**, which is what the
  model uses. The caption is a typo.
- **Table 1's asunaprevir panel is headed "Covariate (n = 256)".** Every count in
  that panel sums to **265** (91 + 174 male/female; 22 + 243 cirrhosis;
  43 + 222 formulation) and the Results text states 265 subjects, so the header
  is a digit transposition.
- **Table 1's asunaprevir AST and ALT maxima appear transposed** relative to the
  daclatasvir panel (595 for AST and 377 for ALT, versus 377 and 595). Only the
  AST median of 52 U/L is load-bearing for the model, and it is consistent across
  the table, the equation text and the forest-plot arithmetic.
- **Table 2's daclatasvir `theta8` confidence interval is printed as
  "-0.189-0.132"**, which as a range spanning zero would contradict the stated
  P < 0.001 significance; the other negative estimates in the same column use the
  "-a to -b" form. The point estimate -0.122 is what the model uses and is
  unambiguous.
- **The covariate equations are images in the published PDF** and are absent from
  text extraction of the article; they were recovered with `pdftotext -layout`
  and each one is independently corroborated by the fold-changes in structural
  check 5.

**Not extracted.**

- Supplementary Tables 1-4 (trial listing, prespecified covariates, base-model
  and full-model parameters) are not on disk. Only the **final** models are
  packaged, which is the library's standing policy for model-development papers,
  so the gap does not affect what is shipped. The base-model IIV values quoted in
  the Results text (CL/F 39.7 %, V/F 40.7 % for daclatasvir; 51.5 % and 97.4 %
  for asunaprevir) are recorded here for completeness.
- The paper mentions later, larger popPK analyses of both drugs that added
  non-Japanese patients (its references 24 and 30, the latter finding race
  significant on asunaprevir CL/F). Those are separate publications and are not
  part of this extraction.
