---
title: "Zanubrutinib (Ou 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Zanubrutinib (Ou 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Ou_2021_zanubrutinib"))$reference`
* Description: `r rxode2::rxode(readModelDb("Ou_2021_zanubrutinib"))$description`
* Article: <https://doi.org/10.1111/cts.12948> (open access, PMC7993273)
* Supplement (demographics, model-development run log, diagnostic plots):
  <https://www.ebi.ac.uk/europepmc/webservices/rest/PMC7993273/supplementaryFiles>

Zanubrutinib is an oral, irreversible, second-generation Bruton's tyrosine
kinase (BTK) inhibitor. Ou 2021 pooled nine clinical studies -- four in
healthy volunteers and five in patients with B-cell malignancies -- and
described the plasma concentrations with a two-compartment model with
sequential zero-order then first-order absorption: the dose enters a depot
compartment at a constant rate over a duration D1 and is then absorbed into
the central compartment with a first-order rate constant ka (Figure 1). Health
status (healthy volunteer vs patient) and baseline alanine aminotransferase
(ALT) were retained as covariates on apparent clearance. Inter-occasion
variability (IOV) was estimated on CL/F, Vc/F and D1, and the residual error
switches from additive-only to combined additive plus proportional at 5 hours
after the previous dose.

```{r load-model}
mod <- readModelDb("Ou_2021_zanubrutinib")
ui <- rxode2::rxode(mod)
ui

# A cl/vc pair can make rxode2 replace the ODEs with its analytic linear
# compartment solution, which would silently drop dur(depot). Confirm the
# explicit ODE system is the one being solved.
stopifnot(is.null(ui$linCmt) || isFALSE(ui$linCmt))
```

# Population

The analysis dataset held 4,925 zanubrutinib plasma concentrations from 632
subjects across nine studies (Table 1 of Ou 2021; Results, Clinical data):
90 healthy volunteers (14.2%) and 542 patients with B-cell malignancies --
CLL/SLL 135 (21.4%), Waldenstrom's macroglobulinemia 196 (31.0%), mantle cell
lymphoma 70 (11.1%) and other B-cell malignancies 141 (22.3%) (Table S1b).
Median age was 64 years (range 19-90; 43 in healthy volunteers, 66 in
patients), median body weight 75 kg (36-144), 29.7% of subjects were female,
and 67.7% were White, 23.1% Asian and 4.1% Black. Median baseline ALT was
18 IU/L (range 4-197) (Table S1a). Doses ranged from 20 to 320 mg once or
twice daily, most subjects receiving 160 mg twice daily; data at 480 mg were
excluded because exposure was less than dose-proportional. The assay LLOQ was
1 ng/mL, and the 6.05% of samples below it were omitted.

```{r population}
str(ui$population)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Ou_2021_zanubrutinib.R`. The
table below collects them in one place.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` | CL/F = 170 L/h (patient) | Table 2, `exp(theta1)`; Eq. 5 coefficient 5.13 |
| `e_dis_healthy_cl` | log(118 / 170) = -0.365 | Table 2, `exp(theta10)` CL/F in healthy volunteers = 118 L/h; Eq. 5 coefficient 4.77 |
| `e_alt_cl` | -0.189 | Table 2, `theta11`; Eq. 5 term `-0.189 x log(ALT/18)` |
| `lvc` | Vc/F = 112 L | Table 2, `exp(theta2)`; Eq. 6 coefficient 4.72 |
| `lq` | Q/F = 26.5 L/h | Table 2, `exp(theta3)` |
| `lvp` | Vp/F = 345 L | Table 2, `exp(theta4)` |
| `lka` | ka = 0.526 1/h | Table 2, `exp(theta5)` |
| `ld1` | D1 = 1.128 h | Table 2, `exp(theta6)` |
| `etalcl`, `etalvc` | 0.367^2, 0.371^2 | Table 2, IIV CL/F 36.7%, Vc/F 37.1% |
| `etalcl`-`etalvc` covariance | 0.129 | Table 2, `Covariance (CL/F, Vc/F)` |
| `etalq`, `etalvp`, `etald1` | 1.02^2, 0.864^2, 0.623^2 | Table 2, IIV Q/F 102%, Vp/F 86.4%, D1 62.3% |
| `etaiov_cl_*`, `etaiov_vc_*`, `etaiov_d1_*` | 0.286^2, 0.675^2, 0.623^2 | Table 2, IOV CL/F 28.6%, Vc/F 67.5%, D1 62.3% |
| `addSd_early` | 8.69 ng/mL | Table 2, `theta9`, additive error for TFDS < 5 h |
| `addSd_late` | 0.633 ng/mL | Table 2, `theta7`, additive error for TFDS >= 5 h |
| `propSd_late` | 0.449 | Table 2, `theta8`, proportional error 44.9% |
| Two-compartment model, zero-order input into depot then first-order absorption | n/a | Figure 1 (equations and diagram); Results, Base model development |
| Covariate form on CL/F | n/a | Eq. 5 |
| IOV occasions (< 7 days and >= 7 days) | n/a | Results, Base model development |
| Residual-error switch at 5 h after the previous dose | n/a | Results, Base model development; Discussion |

# Typical-value checks against Figure 4

Figure 4 of Ou 2021 is a sensitivity analysis of steady-state exposure after
20 doses of 160 mg twice daily. It prints the typical patient's exposure
(ALT = 18 U/L: AUCss 943 ng*h/mL and Cmax,ss 254 ng/mL) and the percent
change for each retained covariate. Those numbers are typical-value
(deterministic) predictions, so they are reproduced here with `zeroRe()`.
Because Cmax depends on ka and D1 as well as on the disposition parameters,
matching Cmax,ss tests the absorption structure, not only the clearance.

```{r helpers}
TAU <- 12
N_DOSES <- 20
SS_START <- (N_DOSES - 1) * TAU   # time of the 20th dose

# Dose rows carry rate = -2 so that dur(depot) <- d1 is honoured; without it
# the dose is a bolus into the depot.
ss_events <- function(scen, rate = -2, fine = 0.01) {
  doses <- expand.grid(id = scen$id, time = (seq_len(N_DOSES) - 1) * TAU)
  doses$amt <- 160
  doses$evid <- 1L
  doses$rate <- rate
  obs <- expand.grid(id = scen$id, time = seq(SS_START, SS_START + TAU, by = fine))
  obs$amt <- NA_real_
  obs$evid <- 0L
  obs$rate <- NA_real_
  ev <- bind_rows(doses, obs)
  ev$cmt <- ifelse(ev$evid == 1L, "depot", "central")
  ev <- merge(ev, scen, by = "id")
  ev[order(ev$id, ev$time, -ev$evid), ]
}

scenarios <- tibble::tribble(
  ~scenario,             ~DIS_HEALTHY, ~ALT,
  "Patient, ALT 18 U/L",            0,   18,
  "Healthy volunteer",              1,   18,
  "ALT 10 U/L",                     0,   10,
  "ALT 37 U/L",                     0,   37
) |>
  mutate(id = row_number(), OCC = 2)

sim_typ <- rxode2::rxSolve(rxode2::zeroRe(mod), events = ss_events(scenarios),
                           keep = "id", returnType = "data.frame") |>
  filter(!is.na(Cc), time >= SS_START) |>
  left_join(scenarios |> select(id, scenario), by = "id")
```

## Steady-state mass balance

For a linear model the solved AUC over one steady-state dosing interval must
equal `Dose / (CL/F)`. Both sides use the same parameters, so the only
residual is integration and trapezoid error.

```{r mass-balance}
mb <- sim_typ |>
  group_by(scenario) |>
  summarise(cl = first(cl),
            auc = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
            cmax = max(Cc), cmin = min(Cc), .groups = "drop") |>
  mutate(auc_closed = 160 * 1000 / cl, rel_err = auc / auc_closed - 1)

knitr::kable(
  mb |>
    select(scenario, cl, auc_closed, auc, rel_err) |>
    dplyr::rename("Scenario" = scenario, "CL/F (L/h)" = cl,
                  "Dose/(CL/F) (ng*h/mL)" = auc_closed,
                  "Solved AUC0-tau (ng*h/mL)" = auc,
                  "Relative error" = rel_err),
  digits = c(0, 2, 1, 1, 7),
  caption = "Steady-state mass balance for the typical-value scenarios."
)

stopifnot(
  # Same parameters on both sides; the residual is trapezoid error on a
  # 0.01 h grid and the approach to steady state after 20 doses (realised
  # about 2e-7).
  max(abs(mb$rel_err)) < 1e-5
)
```

## Figure 4: covariate sensitivity

```{r figure-4-table}
base <- mb |> filter(scenario == "Patient, ALT 18 U/L")

published_fig4 <- tibble::tribble(
  ~scenario,           ~metric, ~published_pct,
  "Healthy volunteer", "AUCss",           43.7,
  "ALT 10 U/L",        "AUCss",          -10.5,
  "ALT 37 U/L",        "AUCss",           14.6,
  "Healthy volunteer", "Cmax,ss",         26.8,
  "ALT 10 U/L",        "Cmax,ss",        -7.37,
  "ALT 37 U/L",        "Cmax,ss",          9.6
)

fig4 <- mb |>
  transmute(scenario,
            AUCss = 100 * (auc / base$auc - 1),
            `Cmax,ss` = 100 * (cmax / base$cmax - 1)) |>
  pivot_longer(c(AUCss, `Cmax,ss`), names_to = "metric", values_to = "model_pct") |>
  inner_join(published_fig4, by = c("scenario", "metric")) |>
  mutate(abs_diff = abs(model_pct - published_pct))

knitr::kable(
  fig4 |>
    dplyr::rename("Scenario" = scenario, "Metric" = metric,
                  "Model (% change)" = model_pct,
                  "Published (% change)" = published_pct,
                  "|difference| (percentage points)" = abs_diff),
  digits = 2,
  caption = "Percent change in steady-state exposure vs the typical patient: model vs Figure 4 of Ou 2021."
)

cmin_alt <- mb |>
  filter(scenario %in% c("ALT 10 U/L", "ALT 37 U/L")) |>
  mutate(pct = 100 * (cmin / base$cmin - 1))

stopifnot(
  # Typical patient: 941.2 vs 943 printed (0.19%) and 253.7 vs 254 (0.13%).
  # The residual is the paper printing CL/F as 170 rather than its unrounded
  # estimate. 1% keeps headroom and still goes red on a unit or dose error.
  abs(base$auc / 943 - 1) < 0.01,
  abs(base$cmax / 254 - 1) < 0.01,
  # Covariate effects: largest gap is the healthy-volunteer AUC (44.07% from
  # the rounded 170/118 vs 43.7% printed). 0.6 points absorbs that rounding;
  # a sign error, a log10 in place of the natural log, or a wrong reference
  # ALT moves these by several points.
  max(fig4$abs_diff) < 0.6,
  # Results: the ALT effect on Cmin,ss is "moderate (< 30%)".
  all(abs(cmin_alt$pct) < 30)
)
```

```{r figure-4, fig.height = 4}
# Replicates Figure 4 of Ou 2021 (covariate bars only; the population 5th-95th
# percentile bar is compared with the simulated cohort further below).
bars <- mb |>
  transmute(scenario, AUCss = auc, `Cmax,ss` = cmax) |>
  pivot_longer(-scenario, names_to = "metric", values_to = "value") |>
  left_join(base |> transmute(AUCss = auc, `Cmax,ss` = cmax) |>
              pivot_longer(everything(), names_to = "metric", values_to = "base"),
            by = "metric") |>
  mutate(covariate = case_when(
    scenario == "Healthy volunteer" ~ "PAT",
    grepl("ALT", scenario) & scenario != "Patient, ALT 18 U/L" ~ "ALT",
    TRUE ~ NA_character_)) |>
  filter(!is.na(covariate))

ggplot(bars, aes(y = covariate)) +
  geom_segment(aes(x = base, xend = value, yend = covariate), linewidth = 8,
               colour = "steelblue") +
  geom_vline(aes(xintercept = base)) +
  geom_text(aes(x = value, label = scenario), vjust = -1.4, size = 3) +
  facet_wrap(~metric, scales = "free_x") +
  labs(x = "Steady-state exposure (AUC ng*h/mL; Cmax ng/mL)", y = NULL,
       title = "Figure 4 -- covariate sensitivity, 160 mg BID",
       caption = "Replicates Figure 4 of Ou 2021. Vertical line: typical patient, ALT 18 U/L.") +
  theme_bw()
```

## The zero-order input is load-bearing

As a control, the same typical patient is solved with the doses entered as a
bolus into the depot (no `rate = -2`, so `dur(depot)` is ignored). The
bolus leaves AUC unchanged but raises Cmax,ss by about 4%, so the 1% Cmax gate
above would fail on it. The gate therefore checks the absorption structure
and not only CL/F.

```{r bolus-control}
bolus <- rxode2::rxSolve(rxode2::zeroRe(mod),
                         events = ss_events(scenarios[1, ], rate = 0),
                         returnType = "data.frame") |>
  filter(!is.na(Cc), time >= SS_START)
bolus_cmax <- max(bolus$Cc)
bolus_cmax

stopifnot(
  abs(bolus_cmax / 254 - 1) > 0.02,
  abs(base$cmax / 254 - 1) < abs(bolus_cmax / 254 - 1)
)
```

# Virtual cohort

Subject-level data are not public. The cohort below is two arms of 200
patients with B-cell malignancies at the two dosing regimens shown in
Figure 3 of Ou 2021 (160 mg twice daily and 320 mg once daily). Baseline ALT
is drawn log-normally with the patient median of 18 IU/L and mean of
22.1 IU/L (Table S1a). Each subject receives eight days of dosing. Records
before day 7 are occasion 1 and later records are occasion 2, matching the
paper's occasion definition, so the first-dose and steady-state intervals
use independent IOV draws.

```{r cohort}
# set.seed() fixes the covariate draw. rxode2's own eta and residual draws are
# partitioned per solver thread and can differ between machines, so every
# assertion on this cohort below is a centre or robust-quantile bound.
set.seed(20210301)
rxode2::rxSetSeed(20210301)

N_PER_ARM <- 200
alt_sdlog <- sqrt(2 * log(22.1 / 18))

arms <- tibble::tribble(
  ~treatment,    ~dose, ~tau,
  "160 mg BID",    160,   12,
  "320 mg QD",     320,   24
)

cohort <- bind_rows(
  tibble(id = seq_len(N_PER_ARM), arms[1, ]),
  tibble(id = 1000L + seq_len(N_PER_ARM), arms[2, ])
) |>
  mutate(ALT = rlnorm(n(), log(18), alt_sdlog), DIS_HEALTHY = 0)

SS_DAY_START <- 192   # start of the day-9 steady-state interval (occasion 2)

cohort_events <- function(cohort) {
  per_id <- lapply(split(cohort, cohort$id), function(s) {
    dose_t <- seq(0, SS_DAY_START, by = s$tau)
    obs_t <- sort(unique(c(
      seq(0, s$tau, by = 0.25),
      SS_DAY_START + seq(0, s$tau, by = 0.25)
    )))
    bind_rows(
      tibble(id = s$id, time = dose_t, evid = 1L, amt = s$dose, rate = -2,
             cmt = "depot"),
      tibble(id = s$id, time = obs_t, evid = 0L, amt = NA_real_,
             rate = NA_real_, cmt = "central")
    )
  })
  ev <- bind_rows(per_id) |>
    left_join(cohort |> select(id, treatment, ALT, DIS_HEALTHY), by = "id") |>
    mutate(OCC = ifelse(time < 7 * 24, 1, 2)) |>
    arrange(id, time, desc(evid))
  as.data.frame(ev)
}

events <- cohort_events(cohort)
stopifnot(!anyDuplicated(events[, c("id", "time", "evid")]))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(mod, events = events, keep = "treatment",
                       returnType = "data.frame") |>
  filter(!is.na(Cc)) |>
  mutate(treatment = as.character(treatment),
         interval = ifelse(time >= SS_DAY_START, "Steady state (day 9)", "First dose"))

stopifnot(!anyNA(sim$Cc), all(sim$Cc >= -1e-6 * max(sim$Cc)))
```

## Figure 3: concentration vs time after the previous dose

Figure 3 of Ou 2021 is a prediction-corrected VPC by regimen, plotted against
time after the previous dose on a log axis. The panels below show the 2.5th,
50th and 97.5th percentiles of the simulated observations (`sim`, which
carries the time-dependent residual error) at steady state.

```{r figure-3}
vpc <- sim |>
  filter(interval == "Steady state (day 9)") |>
  mutate(tad = round(time - SS_DAY_START, 2)) |>
  group_by(treatment, tad) |>
  summarise(lo = quantile(sim, 0.025), med = median(sim),
            hi = quantile(sim, 0.975), .groups = "drop")

ggplot(vpc, aes(tad, med)) +
  geom_ribbon(aes(ymin = pmax(lo, 0.1), ymax = hi), alpha = 0.25, fill = "purple") +
  geom_line(colour = "red") +
  scale_y_log10(limits = c(0.1, 1e4)) +
  facet_wrap(~treatment, scales = "free_x") +
  labs(x = "Time after previous dose (h)", y = "Zanubrutinib concentration (ng/mL)",
       title = "Figure 3 -- steady-state concentration vs time after dose",
       caption = "Replicates Figure 3 of Ou 2021 (simulated 2.5th, 50th, 97.5th percentiles).") +
  theme_bw()
```

## Residual-error structure

The error model is additive only (SD 8.69 ng/mL) for records less than 5 h
after the previous dose, and combined additive (0.633 ng/mL) plus proportional
(44.9%) from 5 h onward. Standardising each simulated residual by the SD that
applies in its stratum should give a unit standard deviation in both strata.
Swapping the two branches, or dropping the switch, moves one of them far from
1.

```{r residual-check}
# Time after the previous dose: in the first interval the only dose is at 0;
# in the steady-state interval the last dose is at SS_DAY_START. Records that
# coincide with a dose (tfds = 0, and the end of the first interval, where the
# next dose is given) are dropped so no record sits on a dose boundary.
res <- sim |>
  left_join(arms |> select(treatment, tau), by = "treatment") |>
  mutate(tfds = ifelse(interval == "First dose", time, time - SS_DAY_START)) |>
  filter(tfds > 0, !(interval == "First dose" & tfds >= tau)) |>
  mutate(stratum = ifelse(tfds >= 5, "TFDS >= 5 h", "TFDS < 5 h"),
         sd_expected = ifelse(tfds >= 5, sqrt(0.633^2 + (0.449 * Cc)^2), 8.69),
         z = (sim - Cc) / sd_expected)

res_tab <- res |>
  group_by(stratum) |>
  summarise(n = n(), sd_z = sd(z), .groups = "drop")

knitr::kable(
  res_tab |> dplyr::rename("Stratum" = stratum, "n" = n,
                           "SD of standardised residual" = sd_z),
  digits = 3,
  caption = "Standardised residuals by time-after-dose stratum; each should have SD near 1."
)

stopifnot(
  # Thousands of draws per stratum; the SD of a unit-variance sample of this
  # size is within a few percent of 1. Swapping the strata would give an SD
  # of order 10 in one of them.
  all(abs(res_tab$sd_z - 1) < 0.1)
)
```

# PKNCA validation

NCA on the steady-state (day 9) dosing interval of each arm, with the time
axis shifted so the interval starts at 0.

```{r pknca}
conc_ss <- sim |>
  filter(interval == "Steady state (day 9)") |>
  transmute(id, treatment, time = time - SS_DAY_START, Cc)
dose_ss <- cohort |> transmute(id, treatment, time = 0, amt = dose)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(as.data.frame(conc_ss), Cc ~ time | treatment + id,
                   concu = "ng/mL", timeu = "h"),
  PKNCA::PKNCAdose(as.data.frame(dose_ss), amt ~ time | treatment + id,
                   doseu = "mg"),
  intervals = data.frame(
    start = 0, end = c(12, 24),
    cmax = TRUE, tmax = TRUE, auclast = TRUE, cmin = TRUE
  ) |> mutate(treatment = c("160 mg BID", "320 mg QD"))
))
```

## Comparison against published exposures

Ou 2021 reports typical-patient steady-state exposure for 160 mg twice daily
(AUCss 943 ng*h/mL, Cmax,ss 254 ng/mL; Figure 4) and a total daily AUC of
1,886 ng*h/mL. For a log-normal clearance the cohort median AUC equals the
typical value, so the median over the virtual cohort is compared with it. The
paper reports no typical exposures for 320 mg once daily. For a linear model
its daily AUC equals the 160 mg twice-daily daily AUC, which is used as the
reference.

Only AUC is gated. Cmax is not log-linear in the random effects, and Vc/F
carries both IIV (37.1%) and IOV (67.5%), a combined log-scale variance of
0.59. The cohort median Cmax therefore sits below the Cmax of the typical
subject (about 15% lower here). The typical-value Cmax of 254 ng/mL is
already matched exactly by the deterministic check above.

```{r nca-compare}
sim_med <- as.data.frame(nca_res$result) |>
  filter(PPTESTCD %in% c("cmax", "auclast"), !is.na(PPORRES)) |>
  group_by(treatment, PPTESTCD) |>
  summarise(PPORRES = median(PPORRES), .groups = "drop")

published <- tibble::tribble(
  ~treatment,    ~cmax, ~auclast,
  "160 mg BID",    254,      943,
  "320 mg QD",      NA,     1886
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_med,
  reference = published,
  by = "treatment",
  units = c(cmax = "ng/mL", auclast = "ng*h/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated cohort median vs the typical-patient exposures of Ou 2021.",
    "The 320 mg QD AUC0-24 reference is the paper's total daily AUC of",
    "1,886 ng*h/mL. * marks a difference above 20%."
  )
)

auc_med <- sim_med |> filter(PPTESTCD == "auclast")

stopifnot(
  # Centre, not extremes: the cohort median AUC must match the typical value.
  # log(AUC) has an SD of about 0.47 (IIV + IOV on CL/F), so the SE of the
  # median over 200 subjects is about 4%; 15% is well over 3 SE and still goes
  # red on a mis-transcribed CL/F, dose or unit.
  abs(auc_med$PPORRES[auc_med$treatment == "160 mg BID"] / 943 - 1) < 0.15,
  abs(auc_med$PPORRES[auc_med$treatment == "320 mg QD"] / 1886 - 1) < 0.15
)
```

Figure 4 also prints the 5th-95th percentile range of steady-state exposure
across the whole analysis population (AUCss 526-2,114 ng*h/mL; Cmax,ss
152-499 ng/mL at 160 mg twice daily). The paper's range comes from the
individual (post hoc) parameters of the 632 analysed subjects, healthy
volunteers included, and is shrunk toward the typical value. The simulated
range comes from the full IIV and IOV in patients only. The comparison is
shown for context and is not gated.

```{r pop-range}
as.data.frame(nca_res$result) |>
  filter(treatment == "160 mg BID", PPTESTCD %in% c("auclast", "cmax")) |>
  group_by(PPTESTCD) |>
  summarise(p05 = quantile(PPORRES, 0.05), p95 = quantile(PPORRES, 0.95),
            .groups = "drop") |>
  mutate(published_p05 = ifelse(PPTESTCD == "auclast", 526.2, 152),
         published_p95 = ifelse(PPTESTCD == "auclast", 2114, 499)) |>
  dplyr::rename("Parameter" = PPTESTCD, "Simulated 5th" = p05,
                "Simulated 95th" = p95, "Published 5th" = published_p05,
                "Published 95th" = published_p95) |>
  knitr::kable(digits = 1, caption = "5th-95th percentile steady-state exposure, 160 mg BID.")
```

# Assumptions and deviations

* **IIV/IOV variance scale.** Table 2 reports variability as percent CV. The
  variances are encoded as `(CV/100)^2`. The printed (CL/F, Vc/F) covariance
  of 0.129 settles this: with `(CV/100)^2` variances (0.1347, 0.1376) the
  implied correlation is 0.947, whereas `log(1 + CV^2)` variances (0.1261,
  0.1289) would imply a correlation of 1.01, which is not a valid covariance
  matrix.

* **IIV and IOV on D1 are printed identically.** Table 2 gives 62.3%
  (95% CI 57.1-67.1; bootstrap 62.2, 51.9-72.4) for both the IIV and the IOV
  of D1. The two agree down to the confidence interval, which suggests one row
  was duplicated when the table was typeset. No other source prints either
  value, so both are encoded as printed.

* **Occasions.** IOV uses two occasions, records < 7 days and >= 7 days after
  the first dose (Results). The `OCC` covariate must be supplied as 1 or 2
  accordingly; a simulation confined to one occasion may use `OCC = 1`
  throughout.

* **Health-status effect.** Eq. 5 gives patients and healthy volunteers
  separate typical log CL/F values (5.13 and 4.77). The model encodes the
  Table 2 typical values, 170 and 118 L/h, as the patient CL/F plus a
  log-ratio shift for `DIS_HEALTHY = 1`. The rounded values give a 44.1%
  higher AUC in healthy volunteers against 43.7% printed in Figure 4. The
  Results text says the health-status effect acts on "CL/F and Vc/F", but
  Eq. 5, Eq. 6 and the run log (Table S2) place it on CL/F only, and that is
  what is encoded.

* **Residual error parameterisation.** The Table 2 residual-error thetas are
  read as standard deviations (additive in ng/mL, proportional as a
  fraction), combined as `sqrt(add^2 + (prop * Cc)^2)` for records 5 h or more
  after the previous dose. The paper's Eq. 2 is the usual combined
  additive-plus-proportional form.

* **Elimination half-life.** The Results report a "geometric mean elimination
  t1/2 of 3.44 hours (CV 40.0%)" without saying how it was derived. For the
  typical parameters the model's terminal half-life is 10.5 h and its
  distribution half-life is 0.39 h. Log-linear fits to the single-dose typical
  profile give 1.9 h over 4-12 h and 4.9 h over 6-24 h. The 3.44 h figure is
  therefore not used as a validation target.

* **Bioavailability.** Only oral data were analysed, so all parameters are
  apparent (CL/F, Vc/F, Q/F, Vp/F). No `f(depot)` term is encoded, and the
  food, CYP3A-interaction and acid-reducing-agent studies enter at face value
  (acid-reducing agents were screened and not retained).

* **Virtual cohort.** Patients only, with baseline ALT drawn log-normally to
  match the Table S1a patient median (18 IU/L) and mean (22.1 IU/L). Other
  demographics are not needed because no other covariate is in the final
  model.
