---
title: "Pazopanib mouse TGI (Ouerdani 2015)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pazopanib mouse TGI (Ouerdani 2015)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Ouerdani A, Struemper H, Suttle AB, Ouellet D, Ribba B. Preclinical modeling of tumor growth and angiogenesis inhibition to describe pazopanib clinical effects in renal cell carcinoma. *CPT Pharmacometrics Syst Pharmacol.* 2015;4(11):660-668. doi:[10.1002/psp4.12001](https://doi.org/10.1002/psp4.12001).
* The on-disk PDF is the corrected version of the article (revised online 2015-11-12); the paper's first-page footnote states that Table 1 was replaced after the initial publication. All parameter values used here come from the corrected Table 1.

This vignette validates the preclinical (mouse, CAKI-2 xenograft) fit of the semi-mechanistic tumour-growth and angiogenesis-inhibition (TGI) model. The paired clinical fit lives in `Ouerdani_2015_pazopanib` and is covered by a separate vignette.

# Population

Female CB-17 SCID mice, 8-10 weeks old, were inoculated subcutaneously with RCC CAKI-2 tumour cells. Once tumours had grown to 100-250 mm^3, mice were randomised into four dose groups of 8 animals each (vehicle, 10, 30, or 100 mg/kg pazopanib once daily by oral gavage for 24 days), giving 32 animals total. Tumour volumes were measured twice weekly by handheld caliper and calculated as `(length * width^2) / 2`. Source: Ouerdani 2015 Methods, preclinical-data section.

The same information is available programmatically via `readModelDb("Ouerdani_2015_pazopanib_mouse")$population`.

# Source trace

Equations from Ouerdani 2015 Methods Equation 1 (preclinical fit with `n = 1`); parameter values from the corrected Table 1, preclinical column. The model has no PK ODE -- drug exposure enters as the per-dose-group covariate `AUC_PAZO`, sourced from the FDA Pharmacology Review for pazopanib NDA 022465 (cited in the paper).

| Equation / parameter           | Value                | Source location                                                              |
|--------------------------------|----------------------|------------------------------------------------------------------------------|
| `d/dt(tumorSize)`              | n/a                  | Ouerdani 2015 Methods Equation 1                                              |
| `d/dt(carryingCapacity)`       | n/a                  | Ouerdani 2015 Methods Equation 1                                              |
| `a = a0 * AUC_PAZO^b_a`        | n/a                  | Ouerdani 2015 Methods Equation 2                                              |
| `c = c0 * AUC_PAZO^b_c`        | n/a                  | Ouerdani 2015 Methods Equation 3                                              |
| `lk_tumor`                     | `log(0.166)`         | Table 1 preclinical k = 0.166 (RSE 24%)                                       |
| `lk_cap` (b)                   | `log(0.0183)`        | Table 1 preclinical b = 0.0183 (RSE 58%); IIV fixed to 0                      |
| `lk_aa0` (c0)                  | `log(0.007)`         | Table 1 preclinical c = 0.007 (RSE 29%)                                       |
| `lk_cyto0` (a0)                | `log(0.251)`         | Table 1 preclinical a = 0.251 (RSE 13%)                                       |
| `lk_res` (d)                   | `log(0.196)`         | Table 1 preclinical d = 0.196 (RSE 26%)                                       |
| `lK0`                          | `log(543)`           | Table 1 preclinical K0 = 543 (RSE 15%)                                        |
| `e_auc_pazo_k_aa` (b_c)        | `0.332`              | Table 1 preclinical b_c = 0.332 (RSE 15%)                                     |
| `e_auc_pazo_k_cyto` (b_a)      | `fixed(0)`           | Paper text: ba initially estimated at 0.0002 then fixed to 0                  |
| `n`                            | `fixed(1)`           | Methods + Results: best preclinical fit at n = 1 (profile-tested 0.5-3)       |
| `etalk_tumor`                  | `0.2809` (= 0.53^2)  | Table 1 preclinical IIV on k = 53% (RSE 108%)                                 |
| `etalk_aa0`                    | `0.0361` (= 0.19^2)  | Table 1 preclinical IIV on c0 = 19% (RSE 127%)                                |
| `etalk_cyto0`                  | `0.0576` (= 0.24^2)  | Table 1 preclinical IIV on a0 = 24% (RSE 96%)                                 |
| `etalk_res`                    | `0.1764` (= 0.42^2)  | Table 1 preclinical IIV on d = 42% (RSE 103%)                                 |
| `etalK0`                       | `0.1296` (= 0.36^2)  | Table 1 preclinical IIV on K0 = 36% (RSE 77%)                                 |
| `propSd`                       | `0.14`               | Table 1 preclinical e1 = 14% (RSE 23%)                                        |
| `addSd`                        | `3`                  | Table 1 preclinical e2 = 3 mm^3 (RSE 17%)                                     |
| `AUC_PAZO` for 10 mg/kg/day    | `220.2 ug*h/mL`      | Ouerdani 2015 Methods, citing FDA Pharmacology Review NDA 022465              |
| `AUC_PAZO` for 30 mg/kg/day    | `656.8 ug*h/mL`      | Same                                                                          |
| `AUC_PAZO` for 100 mg/kg/day   | `1140.8 ug*h/mL`     | Same                                                                          |

# Virtual cohort

The CAKI-2 xenograft study used 8 mice per group across four groups (vehicle, 10, 30, 100 mg/kg pazopanib). The virtual cohort below reproduces that allocation. Per-mouse baseline tumour volume (`TUM_VOL`) is drawn from a uniform distribution over the paper's randomisation window (100-250 mm^3); drug exposure (`AUC_PAZO`) is held constant per dose group as reported in the paper.

```{r cohort}
set.seed(2015)

n_per_group <- 8L
dose_groups <- tibble::tribble(
  ~dose_mgkg, ~AUC_PAZO,
  0,             0,
  10,          220.2,
  30,          656.8,
  100,        1140.8
)

obs_times <- seq(0, 24, by = 0.5)  # twice-weekly measurements = ~every 3-4 days

make_cohort <- function(dose_mgkg, AUC_PAZO, id_offset) {
  ids <- id_offset + seq_len(n_per_group)
  per_subject <- tibble(
    id       = ids,
    TUM_VOL  = runif(n_per_group, min = 100, max = 250),
    AUC_PAZO = AUC_PAZO,
    dose_label = paste0(dose_mgkg, " mg/kg")
  )
  per_subject |>
    tidyr::crossing(time = obs_times) |>
    mutate(evid = 0L, amt = 0)
}

events_mouse <- bind_rows(
  make_cohort(0,    0,     id_offset =   0L),
  make_cohort(10,   220.2, id_offset =   n_per_group),
  make_cohort(30,   656.8, id_offset = 2*n_per_group),
  make_cohort(100,  1140.8, id_offset = 3*n_per_group)
) |>
  mutate(dose_label = factor(
    dose_label,
    levels = c("0 mg/kg", "10 mg/kg", "30 mg/kg", "100 mg/kg")
  ))

stopifnot(!anyDuplicated(unique(events_mouse[, c("id", "time", "evid")])))
nrow(events_mouse); n_distinct(events_mouse$id)
```

# Typical-value simulation (Figure 4 left panel)

A typical-value simulation (`zeroRe()`: no IIV, no residual error) reproduces the mechanistic shape Ouerdani 2015 highlights in Figure 4 (left panel): tumour volume rises monotonically in vehicle animals (pure logistic growth), declines monotonically at low pazopanib doses (cytotoxic effect dominates), and shows an initial decrease followed by limited regrowth at higher doses (the "dual-action" pattern as the antiangiogenic carrying-capacity reduction kicks in).

```{r figure-4-mouse}
mod_mouse <- readModelDb("Ouerdani_2015_pazopanib_mouse")
mod_mouse_typ <- mod_mouse |> rxode2::zeroRe()

sim_typ <- rxode2::rxSolve(
  mod_mouse_typ,
  events = events_mouse,
  keep   = c("dose_label")
) |>
  as.data.frame()

# Population-typical median per dose group
typ_summary <- sim_typ |>
  group_by(dose_label, time) |>
  summarise(median_P = median(tumorSize),
            median_K = median(carryingCapacity),
            .groups = "drop")

ggplot(typ_summary, aes(time, median_P, colour = dose_label)) +
  geom_line(linewidth = 1) +
  labs(x = "Time (days)", y = "Tumour volume P (mm^3)",
       colour = "Dose group",
       title = "Typical mouse trajectory by dose group",
       caption = "Replicates Figure 4 (left panel) of Ouerdani 2015.") +
  theme_minimal()
```

Plotting the carrying capacity `K` alongside `P` shows the angiogenesis-inhibition mechanism: at zero exposure the carrying capacity grows unimpeded; under treatment, the antiangiogenic effect drives `K` below `P` and forces tumour shrinkage.

```{r p-vs-k-mouse}
typ_long <- typ_summary |>
  pivot_longer(c(median_P, median_K), names_to = "state", values_to = "value") |>
  mutate(state = recode(state, median_P = "P (tumour)", median_K = "K (capacity)"))

ggplot(typ_long, aes(time, value, colour = state, linetype = state)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ dose_label, ncol = 2, scales = "free_y") +
  labs(x = "Time (days)", y = "mm^3",
       colour = NULL, linetype = NULL,
       title = "Typical P and K trajectories by dose group",
       caption = "Replicates the mechanistic interpretation of Figure 4 of Ouerdani 2015.") +
  theme_minimal()
```

# Stochastic simulation with IIV

Including IIV (`etalk_tumor ~ 0.2809`, `etalk_aa0 ~ 0.0361`, `etalk_cyto0 ~ 0.0576`, `etalk_res ~ 0.1764`, `etalK0 ~ 0.1296`) and the combined additive + proportional residual error gives a stochastic VPC analogous to Figure 2 of the paper (preclinical VPC stratified by dose).

```{r vpc-mouse}
sim_iiv <- rxode2::rxSolve(
  mod_mouse,
  events = events_mouse,
  keep   = c("dose_label")
) |>
  as.data.frame()

vpc_summary <- sim_iiv |>
  group_by(dose_label, time) |>
  summarise(
    Q05 = quantile(sim, 0.05, na.rm = TRUE),
    Q50 = quantile(sim, 0.50, na.rm = TRUE),
    Q95 = quantile(sim, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(vpc_summary, aes(time, Q50, colour = dose_label, fill = dose_label)) +
  geom_ribbon(aes(ymin = pmax(Q05, 0), ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 1) +
  facet_wrap(~ dose_label, ncol = 2, scales = "free_y") +
  labs(x = "Time (days)", y = "Tumour volume P (mm^3)",
       title = "Stochastic VPC: 5/50/95 percentile bands by dose group",
       caption = "Replicates Figure 2 of Ouerdani 2015 (preclinical VPC). Per-mouse residual error applied; 5th-percentile band clipped at 0.") +
  theme_minimal() +
  guides(colour = "none", fill = "none")
```

The 8 animals per group in the source dataset produce empirical 5th and 95th percentiles equal to the smallest and largest trajectory respectively; the paper itself notes that "VPC should be interpreted carefully, as the number of mice for each group is small." The bands above use a 32-mouse virtual cohort (matching the source N) and so display the same wide spread.

# Mechanistic sanity checks

This is a tumour-growth ODE model rather than a PK model, so PKNCA-based validation is not the right target (there is no drug-concentration / dose / AUC integration to perform). The mechanistic checks below catch the failure modes that would actually break this class of model.

## 1. Vehicle = pure logistic growth

At `AUC_PAZO = 0`, both drug-effect rates `a` and `c` collapse to zero via the model's `if (AUC_PAZO > 0)` gate, and the ODE reduces to the classical Verhulst-Pearl logistic with an antiangiogenesis-driven capacity growth term. The simulated vehicle group should therefore show monotonic tumour expansion across the 24-day window.

```{r veh-check-mouse}
veh_typ <- sim_typ |> filter(dose_label == "0 mg/kg")

# Monotonic non-decreasing tumour volume in vehicle (typical-value, no noise).
stopifnot(all(diff(veh_typ$median_P) > 0))

knitr::kable(
  veh_typ |> filter(time %in% c(0, 6, 12, 18, 24)),
  digits = 3,
  caption = "Typical vehicle-arm trajectory (no drug effect, pure logistic growth)."
)
```

## 2. Drug-effect rates respect the AUC > 0 gate

For any individual with `AUC_PAZO > 0` the per-individual rates `cyto_rate` and `aa_rate` are positive; for vehicle animals they are exactly zero. The gate is what keeps the cytotoxic rate `a = a0 * AUC^0 = a0` from spuriously acting on vehicle animals in the mouse fit (where the exponent `b_a` is fixed to 0 and NONMEM-style `0^0 = 1` would otherwise leave `a` switched on).

```{r gate-check-mouse}
gate_check <- sim_iiv |>
  group_by(dose_label) |>
  summarise(any_cyto = any(cyto_rate > 0),
            any_aa   = any(aa_rate > 0),
            .groups = "drop")

knitr::kable(
  gate_check,
  caption = "Drug-effect rates are zero in vehicle and positive in every drug arm."
)
stopifnot(identical(gate_check$any_cyto, c(FALSE, TRUE, TRUE, TRUE)))
stopifnot(identical(gate_check$any_aa,   c(FALSE, TRUE, TRUE, TRUE)))
```

## 3. Dimensional analysis of the ODE

| Term                                | Units                        | Reduces to     |
|-------------------------------------|------------------------------|----------------|
| `k_tumor * tumorSize`               | `1/day * mm^3`               | `mm^3 / day`   |
| `(1 - tumorSize / carryingCapacity)`| unitless                     | unitless       |
| `cyto_rate * exp(-k_res*t) * tumorSize` | `1/day * unitless * mm^3` | `mm^3 / day`   |
| `k_cap * tumorSize^n`               | `1/day * (mm^3)^1`           | `mm^3 / day` (mouse n = 1) |
| `aa_rate * carryingCapacity`        | `1/day * mm^3`               | `mm^3 / day`   |

Both ODE right-hand sides come out as `mm^3 / day`, consistent with `d/dt(state)` where `state` is in `mm^3` and `t` is in days. The mouse fit's `n = 1` makes `k_cap`'s units cleanly `1/day`; in the paired clinical model `n = 0.5` so `k_cap`'s units carry an implicit `mm^(0.5)` scaling that absorbs the source paper's preclinical-to-clinical re-parameterisation.

# Assumptions and deviations

* **`checkModelConventions()` deviations (intentional).** Running `nlmixr2lib::checkModelConventions("Ouerdani_2015_pazopanib_mouse")` flags four warnings (no errors): (1) the `tumorSize` compartment is not on the canonical-compartment list; (2) the `carryingCapacity` compartment is not on the canonical-compartment list; (3) the single-output observation variable should be `Cc`; (4) `units$concentration` does not contain `/` (mass / volume) and `units$dosing` is `n/a`. All four are intrinsic to a tumour-volume-dynamics model with no drug-concentration / dosing ODE. The same deviations apply to every other tumour-size-dynamics model in the package (e.g., `tgi_sat_logistic`, `oncology_xenograft_simeoni_2004`, `Zecchin_2016_tumorovarian`). Tumour volume in mm^3 is not a drug concentration; `Cc` is reserved for plasma drug concentrations and would be misleading here. The deviations are intentional.
* **`AUC_PAZO > 0` gate added to handle vehicle animals.** Ouerdani 2015 fits the mouse model with `b_a` (AUC exponent on the cytotoxic effect) fixed to 0, which mathematically gives `a = a0` at any positive AUC and, under the standard NONMEM convention `0^0 = 1`, also `a = a0` at AUC = 0. That keeps the cytotoxic rate switched on in vehicle animals, contradicting the paper's vehicle-arm Figure 1 (pure logistic growth, no shrinkage). The nlmixr2lib model file gates both drug-effect rates on `AUC_PAZO > 0` so that vehicle animals see `a = 0` and `c = 0`, recovering the published vehicle-arm trajectory. The paper itself does not publish the source NONMEM control stream, so the exact form of the source author's vehicle-handling cannot be confirmed; the gate is the simplest interpretation consistent with the reported Figure 1 vehicle trajectory and Equation 1 / Equations 2-3.
* **Per-mouse baseline tumour volume drawn uniformly from 100-250 mm^3.** The paper sets P0 per-animal from the observed caliper measurement at randomisation but does not report individual values. The virtual cohort here draws `TUM_VOL` from `Uniform(100, 250)`; switching to a triangular or empirical distribution would shift the per-group dispersion but not the typical-value trajectory.
* **`AUC_PAZO` per dose group is treated as time-fixed.** The paper draws each dose group's mean AUC from the FDA Pharmacology Review for pazopanib NDA 022465; per-animal AUC variability around the group mean is not retained in the model.
* **PKNCA validation is omitted.** This is a tumour-volume-dynamics model with no drug-concentration / dosing ODE -- there is nothing for PKNCA to integrate. The vignette substitutes the mechanistic-sanity checks above (vehicle = pure growth, gate respected, dimensional analysis), matching the validation pattern other tumour-volume models in the package use (e.g., `Zecchin_2016_tumorovarian`).
