---
title: "Pazopanib (Ozbey 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pazopanib (Ozbey 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Ozbey_2021_pazopanib"))$reference`
* Description: `r rxode2::rxode(readModelDb("Ozbey_2021_pazopanib"))$description`
* Article: <https://doi.org/10.3390/ph14090927> (open access)

Ozbey et al. built a one-compartment population PK model of oral pazopanib from
therapeutic-drug-monitoring (TDM) samples and full profiles from a phase I/II
trial. The aim was to estimate each patient's apparent clearance from a single
sample taken at any time after the dose, and then the AUC as
`AUC = Dose / (CL/F)`. Comparing those AUCs with the observed troughs gave a
target AUC of 750 mg\*h/L, which matches the established 20.5 mg/L trough
target.

# Population

The model was fitted to 73 patients (Methods 4.1, Table 4):

- 58 patients from routine pazopanib TDM at Gustave Roussy between 2012 and
  2018. These were 55 adults and 3 children treated for soft-tissue sarcoma or
  Ewing sarcoma, contributing 126 samples (1-6 per patient, drawn at a mean of
  24.6 h post-dose).
- 15 adults with glioblastoma from a phase I/II trial of pazopanib plus
  temozolomide (NCT02331498). They had full 0-24 h profiles (0, 0.5, 1, 2, 4,
  6, 8 and 24 h) on day 1 of cycles 1, 2 and 4, giving 36 profiles and 280
  samples.

Pazopanib was given orally at 200-800 mg once daily; 52 patients were dosed
fasted, 15 fed and 6 unknown. In Table 4, age ranges from 3 to 87 years (median
50.5), 44 patients are male and 29 female, and aspartate aminotransferase
(ASAT) has a median of 36.5 UI/L (range 22.0-233). Missing covariate values were
imputed with the patient's own median, or failing that the population median.
That imputation explains why the ASAT interquartile range (35.4-37.0) is so
narrow. Race, body weight and height are not reported.

The same information is available programmatically via
`readModelDb("Ozbey_2021_pazopanib")()$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Ozbey_2021_pazopanib.R`. The table
below collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| One-compartment, first-order absorption, linear elimination | n/a | Results paragraph 1; Methods 4.3 |
| `lka` (ka) | log(0.976) 1/h | Table 1 row `ka`, RSE 12.3% |
| `lvc` (V/F at ASAT 36.5 UI/L) | log(22.3) L | Table 1 row `V/F`, RSE 9.25% |
| `lcl` (CL/F) | log(0.458) L/h | Table 1 row `Cl/F`, RSE 9.73% |
| `e_ast_vc` | -0.838 | Table 1 covariate column; Methods Equation 4 |
| ASAT centring value | 36.5 UI/L | Methods text below Equation 4; Table 4 median |
| `etalka` | 0.211^2 = 0.044521 | Table 1 IIV column (SD scale, see below) |
| `etalvc` | 0.248^2 = 0.061504 | Table 1 IIV column (SD scale) |
| `etalcl` | 0.714^2 = 0.509796 | Table 1 IIV column (SD scale) |
| `etaiov_vc_1..6` | 0.384^2 = 0.147456 | Table 1 IOV column (SD scale); Methods Equation 1 |
| `etaiov_cl_1..6` | 0.371^2 = 0.137641 | Table 1 IOV column (SD scale); Methods Equation 2 |
| `addSd` | 4.9 mg/L | Table 1 row `Constant residual error`, RSE 8.07% |
| `propSd` | 0.05 | Table 1 row `Proportional residual error`, RSE 26.2% |
| `vc <- exp(lvc + etalvc + iov_vc) * (AST / 36.5)^e_ast_vc` | n/a | Methods Equations 1 and 4 |
| `cl <- exp(lcl + etalcl + iov_cl)` | n/a | Methods Equation 2 |
| `Cc ~ add(addSd) + prop(propSd) + combined1()` | n/a | Methods 4.3 ('combined model'); Monolix 2018 combined1 form |

## Scale of the variability terms

The Table 1 footnote says that "inter-individual variability (IIV) is expressed
in omega^2". The printed values are nonetheless Monolix standard deviations
(omega), for two reasons:

- The Results text reads the V/F entry 0.248 directly as a coefficient of
  variation: ASAT "helped to decrease the inter-individual variability of V/F
  from 39.8% to 24.8%".
- Figure 4 plots the population distribution of each individual parameter. For
  a log-normal with median `m` and log-scale SD `s`, the mode is
  `m * exp(-s^2)` and the density at the mode is
  `exp(s^2 / 2) / (m * s * sqrt(2 * pi))`.

The ka panel is the cleanest test, because ka carries neither IOV nor a
covariate. The curve peaks at a density of about 2 near ka = 0.9 per hour.
The V/F and CL/F panels include the IOV term, which Monolix adds to the
occasion-level parameter. Their curves have modes near 18 L and 0.2-0.25 L/h.
All three were read off Figure 4 by the maintainers.

```{r omega-scale}
m <- c(ka = 0.976, vc = 22.3, cl = 0.458)
iiv <- c(ka = 0.211, vc = 0.248, cl = 0.714)
iov <- c(ka = 0, vc = 0.384, cl = 0.371)

lnorm_mode <- function(m, s) m * exp(-s^2)
lnorm_peak <- function(m, s) exp(s^2 / 2) / (m * s * sqrt(2 * pi))

s_sd <- sqrt(iiv^2 + iov^2)   # table values read as SDs
s_var <- sqrt(iiv + iov)      # table values read as variances

omega_check <- tibble(
  parameter = c("ka (1/h)", "V/F (L)", "CL/F (L/h)"),
  figure4_mode = c("~0.9", "~18", "0.20-0.25"),
  mode_sd_reading = signif(lnorm_mode(m, s_sd), 3),
  mode_variance_reading = signif(lnorm_mode(m, s_var), 3),
  peak_sd_reading = signif(lnorm_peak(m, s_sd), 3),
  peak_variance_reading = signif(lnorm_peak(m, s_var), 3)
)

omega_check |>
  dplyr::rename(
    "Parameter" = parameter,
    "Figure 4 mode (digitised)" = figure4_mode,
    "Mode, SD reading" = mode_sd_reading,
    "Mode, variance reading" = mode_variance_reading,
    "Peak density, SD reading" = peak_sd_reading,
    "Peak density, variance reading" = peak_variance_reading
  ) |>
  knitr::kable(caption = "Figure 4 parameter distributions under the two readings of Table 1.")

# Deterministic arithmetic. The ka curve in Figure 4 peaks at a density of
# about 2 (tallest histogram bars 2.2-2.3). The SD reading gives 1.98 and the
# variance reading 0.99, so it discriminates by a factor of 2. The
# V/F mode (Figure 4: about 18 L) separates the readings the same way:
# 18.1 L versus 11.9 L.
stopifnot(
  abs(lnorm_peak(m[["ka"]], s_sd[["ka"]]) - 2) < 0.3,
  abs(lnorm_peak(m[["ka"]], s_var[["ka"]]) - 2) > 0.8,
  abs(lnorm_mode(m[["vc"]], s_sd[["vc"]]) - 18) < 2,
  abs(lnorm_mode(m[["vc"]], s_var[["vc"]]) - 18) > 5
)
```

Only the SD reading reproduces Figure 4, so the packaged model uses the squared
Table 1 values as the variances.

# Virtual cohort

Individual data are not public. The virtual cohort has 150 subjects in each of
four once-daily dose groups spanning the paper's 200-800 mg range. ASAT is
drawn by interpolating the empirical quantile function through the Table 4
minimum, quartiles and maximum, so the virtual ASAT distribution reproduces
Table 4. Every subject is simulated at steady state on a single occasion
(`OCC = 1`).

```{r cohort}
# rxode2's simulation RNG is partitioned per solver thread, so this cohort is
# reproducible on a given machine but not across thread counts. Every
# assertion on the cohort below is written to hold for any draw the model can
# produce.
rxode2::rxSetSeed(20210915)
set.seed(20210915)

ast_q <- c(22.0, 35.4, 36.5, 37.0, 233) # Table 4: min, Q1, median, Q3, max
draw_ast <- function(n) stats::approx(c(0, 0.25, 0.5, 0.75, 1), ast_q, xout = stats::runif(n))$y

obs_times <- c(0, 0.5, 1, 2, 3, 4, 6, 8, 12, 16, 20, 24)

make_cohort <- function(n, dose, id_offset = 0L) {
  subj <- tibble(
    id = id_offset + seq_len(n),
    AST = draw_ast(n),
    OCC = 1L,
    treatment = paste(dose, "mg QD")
  )
  dosing <- subj |>
    mutate(time = 0, evid = 1L, amt = dose, ii = 24, ss = 1L, cmt = "depot")
  obs <- subj |>
    tidyr::crossing(time = obs_times) |>
    mutate(evid = 0L, amt = 0, ii = 0, ss = 0L, cmt = "central")
  bind_rows(dosing, obs) |>
    arrange(id, time, desc(evid))
}

doses <- c(200, 400, 600, 800)
events <- bind_rows(lapply(seq_along(doses), function(i) {
  make_cohort(150L, doses[i], id_offset = (i - 1L) * 150L)
}))
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod <- readModelDb("Ozbey_2021_pazopanib")

sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep = c("treatment", "AST"),
  maxsteps = 1e6
) |>
  as.data.frame()

mod_typical <- mod |> rxode2::zeroRe()
events_typical <- events |>
  filter(id %in% c(1L, 151L, 301L, 451L)) |>
  mutate(AST = 36.5)
sim_typical <- rxode2::rxSolve(
  mod_typical,
  events = events_typical,
  keep = c("treatment"),
  maxsteps = 1e6,
  rtol = 1e-10,
  atol = 1e-12,
  ssRtol = 1e-10,
  ssAtol = 1e-12
) |>
  as.data.frame()
```

## Typical-value checks

At the median ASAT the typical half-life is `log(2) * V/F / (CL/F)`. For
steady-state once-daily dosing, the paper computes AUC as `Dose / (CL/F)`.

```{r typical}
t_half <- log(2) * 22.3 / 0.458
t_half

typ <- sim_typical |>
  group_by(treatment) |>
  summarise(
    ctrough = Cc[time == 24],
    ctrough_t0 = Cc[time == 0],
    cmax = max(Cc),
    .groups = "drop"
  )
knitr::kable(typ, digits = 2, caption = "Typical-value steady-state trough and peak (mg/L), ASAT 36.5 UI/L.")

# The pre-dose value at t = 0 of an ss = 1 dose is the steady-state trough,
# so it must equal the 24-h value. A mis-set ss record breaks this.
stopifnot(max(abs(typ$ctrough_t0 / typ$ctrough - 1)) < 1e-4)
```

The typical half-life is `r round(t_half, 1)` h. The paper's Introduction
quotes a label half-life of 30.9 h.

# Replicate published figures

## Figure 2 -- visual predictive check

Figure 2 of Ozbey 2021 is a Monolix VPC of concentration against time after
dose (0-30 h), pooling every dose level and occasion. Its observed median runs
from about 35-40 mg/L near the 2-4 h peak to about 20 mg/L at 24 h. The 90%
prediction band spans roughly 5 to 60-100 mg/L. The simulated band below
includes residual error (`sim`) and pools the four virtual dose groups.

```{r figure-2}
vpc <- sim |>
  group_by(time) |>
  summarise(
    Q05 = quantile(sim, 0.05, na.rm = TRUE),
    Q50 = quantile(sim, 0.50, na.rm = TRUE),
    Q95 = quantile(sim, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(vpc, aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  labs(
    x = "Time after dose (h)",
    y = "Pazopanib concentration (mg/L)",
    title = "Steady-state VPC, 200-800 mg once daily pooled",
    caption = "Replicates Figure 2 of Ozbey 2021 (5th, 50th, 95th percentiles)."
  )
```

The simulated band has the same shape and order of magnitude as Figure 2.
Figure 2 also contains first-dose profiles (cycle 1, day 1) and an unknown mix
of dose levels, so a point-by-point comparison is not possible.

## Figure 4 -- distribution of individual parameters

```{r figure-4}
par_df <- sim |>
  distinct(id, ka, vc, cl) |>
  tidyr::pivot_longer(c(ka, vc, cl), names_to = "parameter", values_to = "value") |>
  mutate(parameter = factor(parameter, c("ka", "vc", "cl"), c("ka (1/h)", "V/F (L)", "CL/F (L/h)")))

ggplot(par_df, aes(value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40) +
  facet_wrap(~parameter, scales = "free") +
  labs(
    x = NULL, y = "Density",
    caption = "Replicates Figure 4 of Ozbey 2021 (distribution of individual parameters)."
  )
```

# PKNCA validation

The typical-value steady-state profiles are run through PKNCA over the 0-24 h
dosing interval. The reference values are the paper's own `Dose / (CL/F)` AUC
formula evaluated at the Table 1 CL/F.

```{r pknca}
nca_input <- function(d) {
  d |>
    filter(!is.na(Cc)) |>
    select(id, time, Cc, treatment)
}

dose_input <- function(ids) {
  events |>
    filter(evid == 1, id %in% ids) |>
    select(id, time, amt, treatment)
}

intervals <- data.frame(
  start = 0, end = 24,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE
)

nca_typ <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(nca_input(sim_typical), Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(dose_input(unique(sim_typical$id)), amt ~ time | treatment + id),
  intervals = intervals
))

published <- tibble(
  treatment = paste(doses, "mg QD"),
  auclast = doses / 0.458
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_typ,
  reference = published,
  by = "treatment",
  units = c(auclast = "mg*h/L", cmax = "mg/L", cmin = "mg/L", tmax = "h"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = "Typical-value steady-state NCA vs the paper's AUC = Dose / (CL/F). * differs by >20%."
)

auc_typ <- as.data.frame(nca_typ$result) |>
  filter(PPTESTCD == "auclast") |>
  left_join(published, by = "treatment")
# Same drawn (zero) parameters on both sides: the only difference is the
# trapezoidal error of the 12-point grid across the absorption peak.
stopifnot(max(abs(auc_typ$PPORRES / auc_typ$auclast - 1)) < 0.02)
```

The stochastic cohort goes through the same PKNCA block, split by dose group:

```{r pknca-cohort}
nca_sim <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(nca_input(sim), Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(dose_input(unique(sim$id)), amt ~ time | treatment + id),
  intervals = intervals
))

as.data.frame(nca_sim$result) |>
  group_by(treatment, PPTESTCD) |>
  summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  dplyr::rename(
    "Dose group" = treatment,
    "AUC0-24 (mg*h/L)" = auclast,
    "Cmax (mg/L)" = cmax,
    "Cmin (mg/L)" = cmin,
    "Tmax (h)" = tmax
  ) |>
  knitr::kable(digits = 1, caption = "Median steady-state NCA of the virtual cohort.")
```

# Target AUC of 750 mg\*h/L

The paper's central result links the 20.5 mg/L trough target to an AUC target.
Among the patients' model-based AUCs (Dose / individual CL/F), a cut-off of
750 mg\*h/L separated troughs above and below 20.5 mg/L with a sensitivity of
99.1% and a specificity of 90.6% (Table 2; the maximum Youden index, 89.8, is
at > 743.9 mg\*h/L). The trough-AUC correlation coefficient was 0.919.

The same analysis is repeated on the virtual cohort. Each subject's AUC is
`Dose / CL_i`, and the trough is the 24-h concentration including residual
error. This is how a TDM trough would be measured.

```{r target-auc}
per_subj <- sim |>
  filter(time == 24) |>
  transmute(
    id, treatment,
    dose = as.numeric(sub(" mg QD", "", treatment)),
    auc = dose / cl,
    ctrough = sim
  )

roc <- tibble(cutoff = seq(300, 1500, by = 5)) |>
  rowwise() |>
  mutate(
    sensitivity = mean(per_subj$auc[per_subj$ctrough > 20.5] > cutoff),
    specificity = mean(per_subj$auc[per_subj$ctrough <= 20.5] <= cutoff),
    youden = sensitivity + specificity - 1
  ) |>
  ungroup()

best <- roc |> slice_max(youden, n = 1, with_ties = FALSE)
at750 <- roc |> filter(cutoff == 750)
r_ctrough_auc <- cor(per_subj$ctrough, per_subj$auc)

tibble(
  quantity = c("Optimal AUC cut-off (mg*h/L)", "Sensitivity at 750 (%)", "Specificity at 750 (%)", "Pearson r (Ctrough, AUC)"),
  simulated = c(best$cutoff, 100 * at750$sensitivity, 100 * at750$specificity, r_ctrough_auc),
  published = c(743.9, 99.1, 90.6, 0.919)
) |>
  dplyr::rename("Quantity" = quantity, "Simulated" = simulated, "Ozbey 2021" = published) |>
  knitr::kable(digits = 3, caption = "Target-AUC analysis: virtual cohort vs Table 2 and Results.")

ggplot(per_subj, aes(auc, ctrough, colour = treatment)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 20.5, linetype = 2) +
  geom_vline(xintercept = 750, linetype = 2) +
  scale_x_log10() +
  labs(
    x = "AUC = Dose / CL/F (mg*h/L)",
    y = "Steady-state trough with residual error (mg/L)",
    colour = NULL
  )
```

```{r target-auc-gate}
# The Youden-optimal cut-off is the argmax of a flat curve, so it is not gated:
# across 13 draws it ranged 705-905 mg*h/L. The gates are instead the
# classification performance AT the published 750 target, which is stable
# across the same draws: sensitivity 0.895-0.944, specificity 0.817-0.906,
# r 0.973-0.987, and a Youden index at most 0.06 below the maximum. A CL/F or
# V/F mis-transcribed by a factor of 2 moves the AUC-trough mapping enough to
# fail one of them.
youden750 <- at750$youden
stopifnot(
  at750$sensitivity > 0.8,
  at750$specificity > 0.7,
  youden750 > best$youden - 0.1,
  r_ctrough_auc > 0.9
)
```

The virtual cohort supports the paper's target: an AUC of 750 mg\*h/L
separates steady-state troughs above and below 20.5 mg/L, and its Youden index
(`r round(youden750, 3)`) is within 0.1 of the best achievable cut-off
(`r best$cutoff` mg\*h/L). The Youden curve is flat around its maximum, so the
optimal cut-off itself changes by roughly 100 mg\*h/L from one virtual cohort to
the next. The simulated sensitivity and specificity at 750 mg\*h/L sit a few
points below the published 99.1% and 90.6%. The main reason is that the
simulated troughs carry the model's residual error (4.9 mg/L + 5%), which blurs
the classification near 20.5 mg/L. For reference, at the typical ka and V/F,
750 mg\*h/L corresponds to a steady-state trough of 24.7 mg/L. The published
threshold came from empirical-Bayes clearances in a cohort whose sampling
times, dose levels and occasions the virtual cohort does not reproduce.

# Assumptions and deviations

- **Scale of IIV and IOV.** Table 1 is footnoted as omega^2, but the Results
  text reads 0.248 as 24.8% and only the SD reading reproduces Figure 4 (see
  "Scale of the variability terms"). The maintainers therefore treat the
  printed IIV and IOV values as standard deviations and square them.
- **Residual-error form.** The paper says only that a "combined model" was
  used, without an equation. The maintainers encoded Monolix 2018R1's default
  combined form, `combined1` (residual SD = a + b * f), with a = 4.9 mg/L and
  b = 0.05. The alternative `combined2` form (SD = sqrt(a^2 + b^2 f^2)) would
  give smaller residual SDs at high concentrations.
- **Occasions.** The paper reports IOV on V/F and CL/F but not how occasions
  were delimited. The trial had three profiled occasions (cycles 1, 2 and 4),
  and the TDM patients had 1-6 samples each. Six occasion slots are encoded,
  and records outside `OCC` 1-6 carry no IOV. The simulations here use
  `OCC = 1`, so the IOV acts as additional between-subject variability.
- **ASAT distribution.** The virtual ASAT values interpolate the Table 4
  quantiles. The narrow interquartile range reflects the paper's median
  imputation of missing values, not the true spread of ASAT in the cohort.
- **Dose groups and food.** The dose mix of the source data is not reported, so
  four equal dose groups (200, 400, 600 and 800 mg) are simulated. Food status
  was recorded but not tested as a covariate, and the model has no food effect.
- **No body-size covariate.** The cohort included three children, aged 3 years
  and up, but no weight effect was retained. The model should not be
  extrapolated to children.
- **Errata.** No correction notice was found for this article (Crossref
  `updated-by` record checked 2026-09-29).
