---
title: "Dexamethasone, erythrocyte-encapsulated (Ozdin 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dexamethasone, erythrocyte-encapsulated (Ozdin 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
mod <- rxode2::rxode(readModelDb("Ozdin_2025_dexamethasone"))
```

* Citation: `r mod$reference`
* Description: `r mod$description`
* Article: <https://doi.org/10.1002/psp4.70103>
* Supplement (Appendix S1, contains Tables S1-S4, Figures S1-S10 and the final
  NONMEM control stream): <https://doi.org/10.1002/psp4.70103>

The EryDex System (EDS) loads dexamethasone sodium phosphate (DSP) into the
patient's own erythrocytes ex vivo. After the loaded cells are infused back,
resident phosphatases dephosphorylate DSP to dexamethasone, which then diffuses
passively across the red-cell membrane over a period of weeks. eDSP therefore
behaves as a sustained-release depot even though it is given intravenously.

The authors evaluated depot, dual-absorption, Michaelis-Menten-absorption and
semi-physiological release structures (supplement Figures S3 and S4) and
retained none of them: adding phase 3 data to the semi-physiological model drove
the inter-compartmental clearance to an implausibly low value and its
between-subject variability could not be estimated, which the authors ascribe to
over-parameterisation. The packaged model is therefore the **plain
two-compartment disposition model with linear elimination** that the paper
selected (Results section 3.3, supplement Figure S2), which the authors
themselves describe as adequate for systemic exposure but not a mechanistic
description of red-cell release.

# Population

The pooled analysis dataset contains 133 subjects and 1719 dexamethasone plasma
concentrations from two studies (Ozdin 2025 Table 1, Table S1):

* **Phase 1** (NCT01925859): 18 healthy adults, single intravenous eDSP infusion
  over approximately 10 min, dense sampling from 15 min out to day 42. Nine
  subjects received the half-low dose (approximately 2.5-5 mg DSP encapsulated)
  and nine the high dose (approximately 14-22 mg).
* **Phase 3 ATTeST** (NCT02770807): 115 patients with ataxia telangiectasia
  (109 pediatric, 6 adult), monthly intravenous eDSP infusions over
  approximately 40 min for one year, sparse sampling (pre-dose, 1 h and 4 h
  post-infusion on day 1, 24 h, and day 15 after the first dose; trough and peak
  around doses 2-5; trough only before dose 6). Fifty-eight patients received
  the low dose (mean 8.3 mg) and 57 the high dose (mean 17.4 mg).

Overall the cohort was `r 47`% female and 57% White / 43% African American or
Black, with a median age of 10.0 years (range 5.00-55.0) and a median weight of
25.0 kg (range 15.1-95.7). By age group there were 24 adults (18 healthy from
phase 1 plus 6 patients from phase 3), 44 children aged 10 to under 17 years,
64 aged 6 to under 10 years, and a **single** subject aged 2 to under 6 years.
The paper repeatedly cautions against extrapolating to the youngest band for
that reason, and no maturation function was incorporated.

The same information is available programmatically:

```{r population}
str(readModelDb("Ozdin_2025_dexamethasone")()$population)
```

# Source trace

Every `ini()` value carries an in-file comment naming its source location in
`inst/modeldb/specificDrugs/Ozdin_2025_dexamethasone.R`. The table collects them
for review. "Table 2" is the final population PK parameter table of the main
article; "control stream" is the `Final PK Model - NONMEM File` listing at the
end of supplement Appendix S1.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` | `log(20.8 * 0.899)` = log(18.70) | Table 2, `Clearance CL (L/h)` = 20.8 (RSE 8%), shifted to the patient reference by the Table 2 `Patient effect on CL` factor 0.899 |
| `lvc` | `log(122)` | Table 2, `Central volume V1 (L)` = 122 (RSE 9%) |
| `lq` | `log(0.358)` | Table 2, `Inter-compartmental clearance Q (L/h)` = 0.358 (RSE 13%) |
| `lvp` | `log(16.8)` | Table 2, `Peripheral volume V2 (L)` = 16.8 (RSE 9%) |
| `e_wt_cl_q` | `fixed(0.75)` | Table 2, `Weight effect on CL` and `Weight effect on Q`, both "0.75 FIXED"; control stream `(WTBL/70)**0.75` |
| `e_wt_vc_vp` | `fixed(1)` | Table 2, `Weight effect on V1` and `Weight effect on V2`, both "1 FIXED"; control stream `(WTBL/70)**1` |
| `e_dis_healthy_cl` | `1 / 0.899` = 1.112 | Table 2, `Patient effect on CL -> * theta_PTNT` = 0.899 (RSE 10%); control stream `THETA(9)**PTNT`. Reciprocal because the canonical column is `DIS_HEALTHY = 1 - PTNT` |
| `etalcl` | `log(1 + 0.414^2)` = 0.1582 | Table 2, `ETA_CL` = 41.4% CV (RSE 51%, shrinkage 26%) |
| `etalvc` | `log(1 + 0.707^2)` = 0.4054 | Table 2, `ETA_V1` = 70.7% CV (RSE 21%, shrinkage 7%) |
| no IIV on Q, V2 | -- | Control stream `$OMEGA 0 FIX` for Q and V2; Results section 3.3 ("Inter-individual variability was introduced only on CL and V1") |
| `propSdHealthy` | 0.181 | Table 2, `Proportional Error (%)-Phase 1` = 18.1% (RSE 10%) |
| `addSdHealthy` | 0.0013 | Table 2, `Additive error (ng/mL)-Phase 1` = 0.0013 (RSE 18%) |
| `propSdPatient` | 0.295 | Table 2, `Proportional Error (%)-Phase 3` = 29.5% (RSE 23%) |
| `addSdPatient` | 24.3 | Table 2, `Additive error (ng/mL)-Phase 3` = 24.3 (RSE 58%) |
| `d/dt(central)`, `d/dt(peripheral1)` | n/a | Supplement Figure S2 (structural schematic); control stream `$SUBROUTINE ADVAN3 TRANS4` |
| `cl`, `vc`, `q`, `vp` covariate forms | n/a | Control stream `$PK` block: `TVCL=THETA(1)*(WTBL/70)**0.75 * THETA(9)**PTNT`, `TVV1=THETA(2)*(WTBL/70)**1`, `TVQ=THETA(3)*(WTBL/70)**0.75`, `TVV2=THETA(4)*(WTBL/70)**1` |
| `Cc ~ prop(propSd) + add(addSd)` | n/a | Control stream `$ERROR`: `W = (PROPn*PROPn*IPRED*IPRED + ADDn*ADDn)**0.5`, `Y = IPRED + W*EPS(1)`, `$SIGMA 1 FIX` |
| `Cc <- 1000 * central / vc` | n/a | Control stream `S1=V1` with dose amounts in ug; re-expressed here with dose in mg (the unit the paper reports doses in) so the 1000 converts mg/L to the ng/mL of Table 2, Table 3 and Tables S1-S4 |

## Two source readings that had to be resolved

**Residual error keyed on study phase.** The control stream branches the
residual error on `PHAS` (1 or 3) while the clearance effect uses `PTNT`
(0 = healthy, 1 = patient). Because every phase 1 subject is a healthy adult and
every phase 3 subject is an ataxia-telangiectasia patient (Methods section 2.1,
Table 1), the two columns partition the pooled dataset identically. Both are
therefore carried on the single canonical covariate `DIS_HEALTHY = 1 - PTNT`.

**Scale of the reported between-subject variability.** Table 2 gives BSV as
"41.4%" and "70.7%" under the heading `Between Subject Variability (CV%)`
without stating whether these are the exact log-normal CV
(`100 * sqrt(exp(omega^2) - 1)`) or the approximation `100 * sqrt(omega^2)`. The
two readings imply materially different variances for V1 (0.4054 versus 0.4998).
The model file encodes the exact-CV reading, on the following evidence:

1. **Decisive.** It reproduces the control stream's `$OMEGA(2,2)` starting
   value: `log(1 + 0.707^2) = 0.4054` versus the listed 0.404706, a 0.2%
   difference, where the alternative reading would be 23% away. The phase 3
   residual-error starting values behave the same way (`PROP2` 0.296021 versus
   the Table 2 final 0.295; `ADD2` 24.2317 versus 24.3), so the control
   stream's initial estimates are evidently a previous run's near-converged
   values, which makes that agreement meaningful rather than coincidental.
2. **Supporting.** It is the convention used throughout `nlmixr2lib`.
3. **Corroborating but not decisive.** It reproduces the simulated Cmax spread
   of Table 3 somewhat better. The check is run explicitly below, where the
   margin between the two readings turns out to be only one to two Monte Carlo
   standard errors at the cohort sizes this vignette is allowed to use -- so it
   points the same way in all three age bands, but would not settle the
   question on its own.

# Model structure and typical-value behaviour

Before simulating cohorts, confirm the deterministic behaviour of the packaged
model against quantities that can be derived in closed form.

```{r typical}
tv <- rxode2::zeroRe(mod)

wt_ref  <- 50.7   # median weight, 10-<17 years (Table 3)
dose_ld <- 8.3    # low dose, mg DSP encapsulated (Methods section 2.6)
tau     <- 720    # monthly dosing interval, h
inf_dur <- 40 / 60  # phase 3 infusion duration, approximately 40 min

# Closed-form clearance and AUCinf for an AT patient at the reference weight.
cl_cf   <- 20.8 * 0.899 * (wt_ref / 70)^0.75
auc_cf  <- dose_ld * 1000 / cl_cf

ev_tv <- rxode2::et(time = 0, amt = dose_ld, dur = inf_dur, cmt = "central") |>
  rxode2::et(sort(unique(c(seq(0, 1, by = 0.05), inf_dur,
                           seq(1.5, 24, by = 0.5),
                           seq(28, tau, by = 2)))), cmt = "central") |>
  as.data.frame() |>
  mutate(WT = wt_ref, DIS_HEALTHY = 0)

sim_tv <- rxode2::rxSolve(tv, ev_tv, returnType = "data.frame")

auc_sim <- sum(diff(sim_tv$time) *
                 (head(sim_tv$Cc, -1) + tail(sim_tv$Cc, -1)) / 2)
```

```{r typical-halflife}
# Analytic terminal (beta) half-life from the eigenvalues of the two-compartment
# system at the reference weight.
vc_r <- 122   * (wt_ref / 70)
q_r  <- 0.358 * (wt_ref / 70)^0.75
vp_r <- 16.8  * (wt_ref / 70)
k10 <- cl_cf / vc_r; k12 <- q_r / vc_r; k21 <- q_r / vp_r
beta_r <- ((k10 + k12 + k21) - sqrt((k10 + k12 + k21)^2 - 4 * k10 * k21)) / 2
alpha_r <- (k10 + k12 + k21) - beta_r

tibble::tibble(
  Quantity = c("CL (L/h)", "AUC0-inf = dose/CL (ng*h/mL)",
               "AUC0-720 simulated (ng*h/mL)", "Cmax simulated (ng/mL)",
               "Tmax simulated (h)", "Distribution t1/2 (h)",
               "Terminal t1/2 (h)", "Terminal t1/2 (days)"),
  Value = c(cl_cf, auc_cf, auc_sim, max(sim_tv$Cc),
            sim_tv$time[which.max(sim_tv$Cc)],
            log(2) / alpha_r, log(2) / beta_r, log(2) / beta_r / 24)
) |>
  knitr::kable(digits = 3,
               caption = "Typical-value behaviour at 50.7 kg, low dose (8.3 mg).")
```

Three checks pass here. `AUC0-720` from the ODE equals the closed-form
`dose / CL` identity, which simultaneously gates the ODE system, the dose
encoding, the `1000 *` unit conversion and the observation grid. The simulated
Tmax lands at the end of the 40-min infusion, matching the paper's statement
that Cmax "is expected to occur at the end of eDSP infusion (approximately
0.67 h)". And the simulated Cmax of about 88 ng/mL is within 1% of the
median 88.1 ng/mL that Ozdin 2025 Table 3 reports for the same age group and
dose.

```{r assert-typical}
# The dose/CL identity must hold to within the trapezoidal error of the grid.
stopifnot(abs(auc_sim / auc_cf - 1) < 0.01)
# Tmax must be the end of infusion, not a solver artefact at t = 0.
stopifnot(abs(sim_tv$time[which.max(sim_tv$Cc)] - inf_dur) < 0.06)
```

# Virtual cohort

Individual data are not public. The cohort below mirrors the Monte Carlo
population of Ozdin 2025 Methods section 2.6: three pediatric age bands crossed
with the three dose levels, all coded as ataxia-telangiectasia patients
(`DIS_HEALTHY = 0`). The paper drew weights from CDC growth charts with a
GAMLSS fit and used 1000 subjects per age band; here each of the nine arms uses
100 subjects, and weights are drawn from a log-normal matched to the median and
CV that Table 3 reports for each band, truncated to the reported minimum and
maximum.

```{r cohort}
set.seed(20250311)

n_arm <- 100

age_bands <- tibble::tribble(
  ~agegrp,   ~wt_median, ~wt_cv, ~wt_min, ~wt_max,
  "2-<6 y",        16.3,  0.208,    10.4,    31.9,
  "6-<10 y",       28.1,  0.264,    16.3,    65.6,
  "10-<17 y",      50.7,  0.213,    24.6,    69.9
)

dose_levels <- tibble::tribble(
  ~dose,          ~amt,
  "H-LD 4.2 mg",   4.2,
  "LD 8.3 mg",     8.3,
  "HD 17.4 mg",   17.4
)

# Draw n truncated log-normal weights matched to a median and CV.
sample_weights <- function(n, wt_median, wt_cv, wt_min, wt_max) {
  sdlog <- sqrt(log(1 + wt_cv^2))
  out <- numeric(0)
  while (length(out) < n) {
    draw <- stats::rlnorm(4 * n, log(wt_median), sdlog)
    out <- c(out, draw[draw >= wt_min & draw <= wt_max])
  }
  out[seq_len(n)]
}

n_doses    <- 6                                  # monthly infusions over 6 months
dose_times <- seq(0, by = tau, length.out = n_doses)

# Dense sampling in the first and sixth intervals (the two the paper tabulates),
# daily sampling in between. 0.05 h spacing in the first hour after each dose
# resolves the end-of-infusion peak, as in Methods section 2.6.
grid_dense  <- sort(unique(c(seq(0, 1, by = 0.05), inf_dur,
                             seq(1.5, 24, by = 0.5), seq(28, tau - 4, by = 4))))
grid_coarse <- seq(0, tau - 24, by = 24)
obs_times   <- sort(unique(c(
  grid_dense,
  dose_times[n_doses] + grid_dense,
  as.numeric(outer(dose_times[2:(n_doses - 1)], grid_coarse, "+"))
)))

make_arm <- function(agegrp, wt_median, wt_cv, wt_min, wt_max,
                     dose, amt, id_offset) {
  wt <- sample_weights(n_arm, wt_median, wt_cv, wt_min, wt_max)
  ev <- rxode2::et(time = dose_times, amt = amt, dur = inf_dur, cmt = "central") |>
    rxode2::et(obs_times, cmt = "central") |>
    rxode2::et(id = seq_len(n_arm)) |>
    as.data.frame()
  ev$WT          <- wt[ev$id]
  ev$id          <- ev$id + id_offset
  ev$DIS_HEALTHY <- 0
  ev$agegrp      <- agegrp
  ev$dose        <- dose
  ev
}

arms <- tidyr::crossing(age_bands, dose_levels) |>
  mutate(
    agegrp = factor(agegrp, levels = age_bands$agegrp),
    dose   = factor(dose,   levels = dose_levels$dose)
  ) |>
  arrange(agegrp, dose) |>
  mutate(id_offset = (row_number() - 1L) * n_arm)

events <- do.call(
  rbind,
  lapply(seq_len(nrow(arms)), function(i) {
    make_arm(as.character(arms$agegrp[i]), arms$wt_median[i], arms$wt_cv[i],
             arms$wt_min[i], arms$wt_max[i], as.character(arms$dose[i]),
             arms$amt[i], arms$id_offset[i])
  })
)
events$agegrp <- factor(events$agegrp, levels = age_bands$agegrp)
events$dose   <- factor(events$dose,   levels = dose_levels$dose)

# IDs must be disjoint across arms or rxSolve silently merges subjects.
stopifnot(length(unique(events$id)) == nrow(arms) * n_arm)
stopifnot(!anyDuplicated(events[events$evid == 0, c("id", "time")]))

events |>
  filter(evid == 1, time == 0) |>
  group_by(agegrp) |>
  summarise(n = n() / nrow(dose_levels),
            `Weight mean (kg)` = mean(WT[dose == dose_levels$dose[1]]),
            `Weight CV%` = 100 * sd(WT[dose == dose_levels$dose[1]]) /
              mean(WT[dose == dose_levels$dose[1]]),
            `Weight median (kg)` = median(WT[dose == dose_levels$dose[1]]),
            .groups = "drop") |>
  rename("Age band" = agegrp, "N per dose arm" = n) |>
  knitr::kable(digits = 1,
               caption = "Simulated weight distributions, for comparison with the Weight rows of Ozdin 2025 Table 3 (means 16.8 / 29.5 / 50.3 kg, CV 20.8 / 26.4 / 21.3%, medians 16.3 / 28.1 / 50.7 kg).")
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(mod, events = events, keep = c("agegrp", "dose", "WT"),
                       atol = 1e-12, rtol = 1e-10)

# Deep-tail round-off guard. By 700 h post-dose the model predicts
# concentrations around 1e-7 ng/mL, seven orders of magnitude below the assay
# LLOQ of 0.5 ng/mL (Methods section 2.1); at that scale the solver can return
# values a few times 1e-11 below zero, which PKNCA would turn into NA and which
# would drop out of a log-scale plot. Clamp them at zero.
sim <- sim |>
  mutate(Cc = pmax(Cc, 0),
         agegrp = factor(as.character(agegrp), levels = age_bands$agegrp),
         dose   = factor(as.character(dose),   levels = dose_levels$dose),
         dose_index = pmin(floor(time / tau) + 1, n_doses),
         t_after_dose = time - (dose_index - 1) * tau)
```

# Replicate published figures

```{r figure-S9}
# Replicates Figure S9 of Ozdin 2025: simulated dexamethasone profiles over the
# month after the FIRST dose, by pediatric age band and dose level.
sim |>
  filter(dose_index == 1, Cc > 0) |>
  group_by(agegrp, dose, t_after_dose) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
            .groups = "drop") |>
  ggplot(aes(t_after_dose / 24, Q50, colour = dose, fill = dose)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line() +
  geom_hline(yintercept = 0.5, linetype = "dashed", linewidth = 0.3) +
  facet_wrap(~agegrp) +
  scale_y_log10() +
  labs(x = "Time after first dose (days)", y = "Dexamethasone (ng/mL)",
       colour = "Dose", fill = "Dose",
       title = "Figure S9 - simulated single-dose profiles",
       caption = paste("Replicates Figure S9 of Ozdin 2025. Median with 5th-95th",
                       "percentile band. Dashed line = assay LLOQ 0.5 ng/mL.")) +
  theme(legend.position = "bottom")
```

```{r figure-S10}
# Replicates Figure S10 of Ozdin 2025: the same profiles over the month
# following the SIXTH monthly dose.
sim |>
  filter(dose_index == n_doses, Cc > 0) |>
  group_by(agegrp, dose, t_after_dose) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
            .groups = "drop") |>
  ggplot(aes(t_after_dose / 24, Q50, colour = dose, fill = dose)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line() +
  geom_hline(yintercept = 0.5, linetype = "dashed", linewidth = 0.3) +
  facet_wrap(~agegrp) +
  scale_y_log10() +
  labs(x = "Time after sixth dose (days)", y = "Dexamethasone (ng/mL)",
       colour = "Dose", fill = "Dose",
       title = "Figure S10 - simulated sixth-dose profiles",
       caption = paste("Replicates Figure S10 of Ozdin 2025. Visually identical",
                       "to Figure S9, i.e. no accumulation.")) +
  theme(legend.position = "bottom")
```

# PKNCA validation

Exposure metrics are computed with PKNCA over two windows: the first monthly
interval (Ozdin 2025 Table 3) and the sixth (Table S3). Both tables report
"AUC over one month", so `auclast` on a `[dose, dose + 720 h]` interval is the
matching parameter.

```{r pknca-setup}
# PKNCA reserves "dose" as one of its standardized column names, so the dose
# group is carried as `dosegrp` from here on.
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, agegrp, dosegrp = dose)

# Guarantee a time-zero record per subject so PKNCA can anchor AUC without
# emitting "Requesting an AUC range starting (0) before the first measurement".
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, agegrp, dosegrp) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, agegrp, dosegrp, time, .keep_all = TRUE) |>
  arrange(id, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | agegrp + dosegrp + id)

dose_df <- events |>
  filter(evid == 1) |>
  select(id, time, amt, agegrp, dosegrp = dose)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | agegrp + dosegrp + id)

intervals <- data.frame(
  start   = c(0, dose_times[n_doses]),
  end     = c(tau, dose_times[n_doses] + tau),
  cmax    = TRUE,
  tmax    = TRUE,
  auclast = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)

nca_long <- as.data.frame(nca_res) |>
  mutate(window = if_else(start == 0, "First dose", "Sixth dose"))
```

`cl.obs` and `vz.obs` are deliberately not requested: the concentrations are in
ng/mL while the dose amounts are in mg, so PKNCA would report them a factor of
1000 away from the model's L/h and L. The closed-form table above covers those
quantities instead.

## Comparison against published exposures - first dose (Table 3)

`ncaComparisonTable()` aggregates the simulated subjects by median, so the
reference column transcribes the **median** values of Ozdin 2025 Table 3 (the
table reports `Mean (CV%)` above `Median [Min, Max]` in each cell). Tmax is not
tabulated; the 0.67 h reference is the end-of-infusion value stated in Results
section 3.5 and the Discussion.

```{r nca-compare-first}
published_first <- tibble::tribble(
  ~agegrp,      ~dosegrp,         ~cmax, ~auclast, ~tmax,
  "2-<6 y",     "H-LD 4.2 mg",   132,      685,  0.67,
  "2-<6 y",     "LD 8.3 mg",     260,     1354,  0.67,
  "2-<6 y",     "HD 17.4 mg",    546,     2838,  0.67,
  "6-<10 y",    "H-LD 4.2 mg",  80.0,      434,  0.67,
  "6-<10 y",    "LD 8.3 mg",     158,      857,  0.67,
  "6-<10 y",    "HD 17.4 mg",    331,     1796,  0.67,
  "10-<17 y",   "H-LD 4.2 mg",  44.6,      292,  0.67,
  "10-<17 y",   "LD 8.3 mg",    88.1,      578,  0.67,
  "10-<17 y",   "HD 17.4 mg",    185,     1211,  0.67
)

cmp_first <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_long |> filter(window == "First dose"),
  reference     = published_first,
  by            = c("agegrp", "dosegrp"),
  params        = c("cmax", "auclast", "tmax"),
  units         = c(cmax = "ng/mL", auclast = "ng*h/mL", tmax = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp_first,
  caption = "Simulated vs. Ozdin 2025 Table 3 medians (single dose). * differs from reference by >20%."
)
```

## Comparison against published exposures - sixth dose (Table S3)

Table S3 of the supplement reports exposures after the sixth monthly dose. Its
numbers are **byte-for-byte identical** to Table 3, which is the paper's
central finding: once-monthly eDSP produces no accumulation. The same reference
values are therefore reused here, and the simulated sixth-dose column is an
independent test of that structural claim.

```{r nca-compare-sixth}
cmp_sixth <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_long |> filter(window == "Sixth dose"),
  reference     = published_first,
  by            = c("agegrp", "dosegrp"),
  params        = c("cmax", "auclast", "tmax"),
  units         = c(cmax = "ng/mL", auclast = "ng*h/mL", tmax = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp_sixth,
  caption = "Simulated vs. Ozdin 2025 Table S3 medians (sixth dose). * differs from reference by >20%."
)
```

## Accumulation ratio

```{r figure-4}
# Replicates Figure 4 of Ozdin 2025: Cmax and AUC after a single dose versus
# after the sixth monthly dose, by age band and dose level.
nca_long |>
  filter(PPTESTCD %in% c("cmax", "auclast")) |>
  mutate(PPTESTCD = recode(PPTESTCD, cmax = "Cmax (ng/mL)",
                           auclast = "AUC (ng*h/mL)")) |>
  ggplot(aes(dosegrp, PPORRES, fill = window)) +
  geom_boxplot(outlier.shape = NA, position = position_dodge(0.8)) +
  facet_grid(PPTESTCD ~ agegrp, scales = "free_y") +
  scale_y_log10() +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "Figure 4 - single dose vs. sixth dose exposure",
       caption = paste("Replicates Figure 4 of Ozdin 2025. Outliers suppressed,",
                       "as in the source figure.")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r accumulation}
accum <- nca_long |>
  filter(PPTESTCD %in% c("cmax", "auclast")) |>
  select(id, agegrp, dosegrp, PPTESTCD, window, PPORRES) |>
  tidyr::pivot_wider(names_from = window, values_from = PPORRES) |>
  mutate(ratio = `Sixth dose` / `First dose`) |>
  group_by(PPTESTCD) |>
  summarise(`Median ratio` = median(ratio),
            `Min ratio` = min(ratio), `Max ratio` = max(ratio),
            .groups = "drop") |>
  rename("Parameter" = PPTESTCD)

knitr::kable(accum, digits = 5,
             caption = "Accumulation ratio (sixth dose / first dose) across all 900 simulated subjects.")

# The paper's core claim: no accumulation with once-monthly dosing.
stopifnot(all(abs(accum$`Median ratio` - 1) < 0.01))
```

The accumulation ratio is 1.000 to within numerical precision, which the
model's parameters make inevitable: the terminal half-life is about 31 h
(computed above), so a 720 h dosing interval is roughly 23 terminal half-lives
and nothing carries over.

# Between-subject-variability scale check

This is the quantitative version of point 2 in the source-trace discussion. The
cohort is re-solved under the alternative reading `omega^2 = CV^2` and the
resulting exposure spread is compared with what Ozdin 2025 Table 3 reports.

Cmax is the only metric with useful discriminating power: it is driven mostly by
V1, where the two readings differ by 23% in variance (0.4054 versus 0.4998).
AUC is driven by CL, where they differ by only 4% (0.1582 versus 0.1714), so AUC
cannot separate them and is not shown.

Two refinements are needed to make the comparison interpretable at 100 subjects
per arm. First, Table 3 reports the **same** Cmax CV% for all three dose levels
within an age band (63.9% throughout the 2-to-under-6 band, 65.9% throughout
6-to-under-10, 71.0% throughout 10-to-under-17), confirming that the paper's own
simulated spread was dose-independent; dose-normalised Cmax can therefore be
pooled across the three dose arms of a band, giving 300 subjects per estimate.
Second, the statistic compared is the **standard deviation of log Cmax** rather
than the raw-scale CV. A raw CV of a heavy-tailed log-normal is a very noisy
estimator at these sample sizes, whereas `sd(log(.))` is estimated with a
Monte Carlo standard error of about `sd / sqrt(2(n - 1))`, which is reported
alongside so the reader can judge whether a gap is meaningful. The published
CV% is put on the same scale via `omega = sqrt(log(1 + CV^2))`.

```{r omega-check}
dose_amt <- stats::setNames(dose_levels$amt, dose_levels$dose)

# sd(log(dose-normalised Cmax)) over the first monthly interval, pooled across
# the three dose arms of each age band, with its Monte Carlo standard error.
log_cmax_spread <- function(solved, reading) {
  solved |>
    filter(time <= tau) |>
    mutate(Cc = pmax(Cc, 0)) |>
    group_by(agegrp, dose, id) |>
    summarise(cmax = max(Cc), .groups = "drop") |>
    mutate(log_cmax_norm = log(cmax / dose_amt[as.character(dose)])) |>
    group_by(agegrp) |>
    summarise(`sd(log Cmax)` = sd(log_cmax_norm),
              `MC SE` = sd(log_cmax_norm) / sqrt(2 * (n() - 1)),
              n = n(), .groups = "drop") |>
    mutate(Reading = reading)
}

# The packaged model (exact-CV reading) reuses the simulation already run above.
spread_exact <- log_cmax_spread(
  sim |> select(id, time, Cc, agegrp, dose),
  "omega^2 = log(1 + CV^2)  [packaged]"
)

# The alternative reading: one extra solve on the identical event table.
mod_alt <- rxode2::ini(mod, etalcl = 0.414^2, etalvc = 0.707^2)
spread_alt <- rxode2::rxSolve(mod_alt, events, keep = c("agegrp", "dose"),
                              atol = 1e-12, rtol = 1e-10) |>
  as.data.frame() |>
  mutate(agegrp = factor(as.character(agegrp), levels = age_bands$agegrp),
         dose   = factor(as.character(dose),   levels = dose_levels$dose)) |>
  select(id, time, Cc, agegrp, dose) |>
  log_cmax_spread("omega^2 = CV^2  [rejected]")

# Table 3 CV% put on the sd(log) scale.
published_spread <- tibble::tibble(
  agegrp = factor(age_bands$agegrp, levels = age_bands$agegrp),
  cv = c(0.639, 0.659, 0.710),
  `sd(log Cmax)` = sqrt(log(1 + cv^2)),
  `MC SE` = NA_real_, n = NA_integer_,
  Reading = "Ozdin 2025 Table 3 (CV% -> sd log)"
) |>
  select(-cv)

omega_check <- bind_rows(published_spread, spread_exact, spread_alt) |>
  mutate(agegrp = factor(as.character(agegrp), levels = age_bands$agegrp)) |>
  arrange(agegrp) |>
  select(Reading, agegrp, `sd(log Cmax)`, `MC SE`, n)

omega_check |>
  rename("Age band" = agegrp, "N pooled" = n) |>
  knitr::kable(digits = 3,
               caption = "Spread of log dose-normalised Cmax under the two readings of Table 2's BSV column, against the published value. Nominal omega_V1 is 0.637 under the packaged reading and 0.707 under the rejected one.")
```

```{r omega-check-summary}
ref <- omega_check |> filter(grepl("^Ozdin", Reading))
omega_check |>
  filter(!grepl("^Ozdin", Reading)) |>
  group_by(Reading) |>
  summarise(
    `Mean deviation from published (sd log units)` =
      mean(`sd(log Cmax)` - ref$`sd(log Cmax)`[match(agegrp, ref$agegrp)]),
    `Mean abs deviation` =
      mean(abs(`sd(log Cmax)` - ref$`sd(log Cmax)`[match(agegrp, ref$agegrp)])),
    `Typical MC SE` = mean(`MC SE`),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 3,
               caption = "Agreement with the published Cmax spread, averaged over the three age bands. Smaller absolute deviation is better; compare it against the Monte Carlo standard error in the last column.")
```

**How much this check proves.** The packaged reading lands within one Monte
Carlo standard error of the published spread in all three age bands (mean
absolute deviation 0.016 against a Monte Carlo SE of about 0.026, and essentially
no systematic bias). The rejected reading is high in every band, by about 0.075
on average, roughly three Monte Carlo standard errors. Per band the gap between
the two readings ranges from about 0.4 to about 4 standard errors, so no single
band is conclusive on its own, but the sign is the same in all three and the
pooled discrepancy is well outside the noise.

This is nevertheless treated as **corroborating rather than decisive**, because
the comparison also depends on the assumed weight distribution (see
Assumptions), which contributes to the Cmax spread independently of `omega`. The
decisive evidence remains the arithmetic agreement between
`log(1 + 0.707^2) = 0.4054` and the control stream's `$OMEGA` starting value of
0.404706 -- a 0.2% match where the alternative would be 23% away -- together
with the `nlmixr2lib` house convention.

# Assumptions and deviations

- **Weight distribution.** Ozdin 2025 drew weights from CDC growth charts using
  a GAMLSS fit; that fitted object is not published. Weights here are truncated
  log-normal draws matched to the median and CV of each age band in Table 3.
  The 10-to-under-17 band is the one place this matters: the published
  distribution is slightly left-skewed (mean 50.3 kg below median 50.7 kg),
  which a log-normal cannot reproduce, so simulated Cmax in that band runs a
  few percent above the published median and its CV% a few points below.
- **Cohort size.** 100 subjects per arm (900 total) rather than the paper's
  1000 per age band, per the `nlmixr2lib` cap of 200 per arm. The comparison
  is against published medians, which are stable at this size.
- **Infusion duration.** Fixed at 40 min for every arm, the phase 3 value
  (Methods section 2.1). The phase 1 infusions were about 10 min, which shifts
  Tmax but has almost no effect on AUC.
- **Dose amounts.** The nominal levels 4.2 / 8.3 / 17.4 mg from Methods section
  2.6 are used for every subject. The actual per-subject encapsulated dose
  varied substantially (Table S2 reports CV 18-47% within dose level, and
  Table S4 shows observed low-dose amounts spanning 0.156-27.3 mg), because the
  mass loaded into the erythrocytes cannot be controlled exactly. Figure 4 of
  the paper labels the low and high doses 8.12 mg and 17.3 mg while Table 3
  labels them 8.3 mg and 17.4 mg; the Table 3 values are used here for
  consistency with the exposures being compared.
- **Molecular-weight conversion.** None is applied. Doses are the mass of DSP
  encapsulated and concentrations are dexamethasone, so a strictly mechanistic
  treatment would scale the dose by `FD = 0.76` (516.4 / 392.464 g/mol). The
  paper introduces `FD` only in the *rejected* semi-physiological model
  (supplement Figure S4 notes); the retained control stream doses the central
  compartment with the DSP amount directly, and reproducing Table 3 confirms
  that reading. The `compartmentData` entries therefore describe amounts of
  dexamethasone on a DSP mass basis.
- **`DIS_HEALTHY` orientation.** The structural typical `lcl` is shifted to the
  ataxia-telangiectasia patient state so that `DIS_HEALTHY = 1` restores the
  paper's printed healthy typical of 20.8 L/h. Simulating healthy adults
  requires setting `DIS_HEALTHY = 1` and supplying adult weights; every arm in
  this vignette is a patient arm.
- **Residual error is not applied in the simulations.** All figures and tables
  show `Cc` (the individual prediction), not `sim` with residual error added.
  The phase 3 additive SD of 24.3 ng/mL is large relative to the tail
  concentrations, so adding it would dominate a semi-log profile plot; the
  published Table 3 and Table S3 exposures are likewise derived from
  model-predicted profiles (Methods section 2.6).

## Errata and unreproducible source claims

- **The Discussion's 20-30 day concentration is off by orders of magnitude.**
  Ozdin 2025 Discussion states that "both LD and HD dose levels of eDSP show
  dexamethasone plasma concentrations of approximately 10^-3 ng/mL after
  20-30 days (corresponding to 500-720 h) from dosing time". With the Table 2
  parameters the model predicts about 1e-6 ng/mL at 500 h and about 1e-7 ng/mL
  at 700 h for the low dose, three to four orders of magnitude lower. The
  arithmetic reason is that the beta-phase intercept is small: at the reference
  weight it is only about 0.3 ng/mL, not the roughly 90 ng/mL of Cmax, so the
  concentration is already far below 10^-3 ng/mL well before day 20. The paper's
  own Figure 2 caption is consistent with the model rather than the Discussion,
  noting that 202 observations had IPRED below an LLOQ of 0.0005 ng/mL. The
  Discussion sentence appears to be an error, and the "sustained release for
  20-30 days" narrative is a property of the *observed* data and of the rejected
  mechanistic release models, not of the retained two-compartment model, whose
  terminal half-life is about 31 h. The paper acknowledges this limitation
  directly: "the simplified two-compartment model does not explicitly capture
  the controlled, likely triphasic release from RBCs".

  ```{r errata-tail}
  sim |>
    filter(dose == "LD 8.3 mg", agegrp == "10-<17 y", dose_index == 1,
           t_after_dose %in% c(24, 240, 480, 720)) |>
    group_by(t_after_dose) |>
    summarise(`Median Cc (ng/mL)` = median(Cc), .groups = "drop") |>
    mutate(`Days after dose` = t_after_dose / 24, .before = 1) |>
    select(-t_after_dose) |>
    knitr::kable(digits = 10,
                 caption = "Model-predicted low-dose tail concentrations, 10-<17 y. Compare with the Discussion's claim of approximately 1e-3 ng/mL at 500-720 h.")
  ```

- **Objective function values.** Results section 3.2 quotes 2652.543 for the
  two-compartment model and 2724.38 for the three-compartment model, while the
  Table 2 note gives 8584.636 for the final model. The first pair evidently
  refers to the phase-1-only structural comparison and the third to the pooled
  final fit; the paper does not say so explicitly. No model parameter depends on
  this.
- **Adult counts in the abstract.** The abstract describes the pooled population
  as "24 healthy adults and 109 AT patients". Table 1 and Results section 3.1
  make clear that the 24 are all *adults* (18 healthy from phase 1 plus 6
  ataxia-telangiectasia patients from phase 3), not 24 healthy subjects. The
  model file uses the Table 1 reading: 18 subjects with `DIS_HEALTHY = 1` and
  115 with `DIS_HEALTHY = 0`.
- **`$THETA` and `$OMEGA` in the supplement control stream are initial
  estimates, not final ones** (CL 19.3324 vs. 20.8; V2 9.94667 vs. 16.8; PTNT
  0.5 vs. 0.899). Every value in the model file comes from Table 2. The
  starting values are used only as corroborating evidence for the BSV scale, as
  described above.
