---
title: "Paracetamol (Padavia 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Paracetamol (Padavia 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Padavia_2024_paracetamol"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1007/s40262-024-01439-3>
* Trial registration: [NCT04459117](https://clinicaltrials.gov/study/NCT04459117) (TREOCAPA)

Padavia 2024 is the first population PK analysis of intravenous paracetamol and
its three measured metabolite classes in *extreme* preterm neonates (23-26
completed weeks' gestational age), a population in which fewer than five
subjects appear in each of the four prior paracetamol PK studies. The paper's
central pharmacological result is not a clearance value but a **pathway split**:
in this population paracetamol is metabolised overwhelmingly by sulphation
(89%), with glucuronidation a distant second (6%) and the toxicologically
relevant oxidative route reduced to under 4%, and the balance between the first
two shifts systematically with gestational age.

# Population

Thirty extreme preterm neonates (17 boys, 13 girls) from eight neonatal
intensive care units in France and Finland, enrolled between November 2020 and
September 2021 in the TREOCAPA phase II dose-escalation trial. Gestational age
at birth spanned 23-26 completed weeks (2, 9, 7 and 12 subjects respectively),
median birth weight 800 g (range 470-920) and median birth length 32.75 cm
(range 28.0-36.5) -- Table 1 of the source. Thirty-one neonates were enrolled
and one was excluded from the PK analysis for having no measurable paracetamol
level.

Treatment began within 12 h of birth and ran for 5 days. Twenty-one subjects
received dose level 1 (20 mg/kg intravenous loading dose then 7.5 mg/kg every
6 h) and nine received dose level 2 (25 mg/kg loading then 10 mg/kg every 6 h);
escalation was stopped at level 2 by the trial's Bayesian continual-reassessment
design because no further efficacy gain was expected. The analysis dataset held
121 paracetamol and 484 metabolite plasma concentrations, a median of 4 samples
per subject (range 2-6).

```{r population}
str(ui$population)
```

# Source trace

Each `ini()` entry in `inst/modeldb/specificDrugs/Padavia_2024_paracetamol.R`
carries an in-file comment naming its origin. The table below collects them.
All parameter values come from **Table 2** of Padavia 2024; the covariate
centring values come from the **Table 2 footnote**, and the structural equations
from **Methods section 2.4**.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` (CL) | 0.0785 L/h | Table 2, parent model (RSE 8.49%; bootstrap 0.0656-0.0920) |
| `e_wt_cl` (beta Cl_W) | 1.81 | Table 2, parent model (RSE 0.302%; bootstrap 1.21-3.40) |
| `lvc` (V1) | 0.124 L | Table 2, parent model (RSE 1.11%; bootstrap 0.0181-0.575) |
| `lq` (Q) | 3.84 L/h | Table 2, parent model (RSE 0.386%; bootstrap 0.616-32.4) |
| `lvp` (V2) | 0.672 L | Table 2, parent model (RSE 10.8%; bootstrap 0.246-0.976) |
| `e_ht_birth_vp` (beta V2_BL) | 6.46 | Table 2, parent model (RSE 24.8%; bootstrap 1.90-10.6) |
| `lclrat_gluc` (t1) | 7.53 | Table 2, parent-metabolite (RSE 19.4%; bootstrap 3.30-9.44) |
| `e_ga_clrat_gluc` (beta t1_GA) | 6.88 | Table 2, parent-metabolite (RSE 1.33%; bootstrap 2.64-12.2) |
| `lclrat_sulf` (t2) | 111 | Table 2, parent-metabolite (RSE 20.1%; bootstrap 36.8-128) |
| `lclrat_cysmer` (t3) | 4.56 | Table 2, parent-metabolite (RSE 2.13%; bootstrap 2.42-5.77) |
| `lcle_gluc` (CL E gluc) | 0.0324 L/h | Table 2, parent-metabolite (RSE 26.3%; bootstrap 0.0168-0.0823) |
| `lcle_sulf` (CL E sulf) | 0.0173 L/h | Table 2, parent-metabolite (RSE 7.66%; bootstrap 0.0132-0.0217) |
| `e_ga_cle_sulf` (beta CLE sulf_GA) | 6.26 | Table 2, parent-metabolite (RSE 1.78%; bootstrap 2.22-10.8) |
| `lcle_cysmer` (CL E ox) | 0.00445 L/h | Table 2, parent-metabolite (RSE 26.5%; bootstrap 0.00232-0.0162) |
| `etalcl` | 0.409 SD -> 0.167281 var | Table 2 omega CL; footnote "omega inter-subject variability expressed as standard deviation" |
| `etalvp` | 0.325 SD -> 0.105625 var | Table 2 omega V2 (same footnote) |
| `etalclrat_gluc` | 0.495 SD -> 0.245025 var | Table 2 omega t1 (RSE 31.3%) |
| `etalclrat_sulf` | 0.924 SD -> 0.853776 var | Table 2 omega t2 (RSE 15.7%) |
| `etalcle_gluc` | 0.835 SD -> 0.697225 var | Table 2 omega CLE gluc (RSE 24.2%) |
| `etalcle_sulf` | 0.278 SD -> 0.077284 var | Table 2 omega CLE sulf (RSE 30.2%) |
| `etalcle_cysmer` | 0.926 SD -> 0.857476 var | Table 2 omega CLE ox (RSE 22.6%) |
| `propSd` | 0.349 | Table 2, proportional error (parent) |
| `propSd_gluc` | 0.443 | Table 2, proportional error for glucuronide |
| `propSd_sulf` | 0.503 | Table 2, proportional error for sulfate |
| `propSd_cysmer` | 0.470 | Table 2, proportional error for oxidative metabolites |
| Covariate form `theta = theta_pop * (cov / Me(cov))^beta` | n/a | Methods 2.4, continuous-covariate equation |
| Centring values: 800 g weight, 33 cm birth length, 25 weeks GA | n/a | Table 2 footnote |
| `f_gluc = t1 / (t1 + t2 + t3 + 1)` and siblings | n/a | Methods 2.4 fraction equations; repeated in the Table 2 footnote |
| `CL_pathway = CL * f_pathway` | n/a | Methods 2.4, "The clearance associated with each elimination pathway is obtained by multiplying total clearance of paracetamol (CL) by the corresponding metabolisation fraction" |
| Metabolite volumes = parent V1 | n/a | Methods 2.4, "The metabolite distribution volumes are not identifiable in parent-metabolite models, so they were fixed to the parent volume" |
| Five-compartment structure | n/a | Results 3.2 and supplemental Figure 1 |
| Exponential IIV `theta_i = theta_pop * exp(eta_i)` | n/a | Methods 2.4 |
| Molar (umol/L) basis | n/a | Methods 2.4, doses and concentrations converted using molecular weights |

## Two published models, one file

Table 2 reports a **parent model** and a **parent-metabolite model**. Methods 2.4
states that "the parent-metabolite model was developed by setting the parameters
of the previously constructed parent model", and Table 2 accordingly repeats
`V1`, `CL`, `beta Cl_W`, `Q`, `V2`, `beta V2_BL`, `omega CL`, `omega V2` and the
parent proportional error in the parent-metabolite block with `-` in both
uncertainty columns. The final parent-metabolite model therefore *contains* the
parent model exactly, so this is one model file (per
`references/replicate-author-structure.md`, a base-plus-final development
sequence extracts as the final model only). Simulating this model and reading
only `Cc` reproduces the parent model; the carried-over parameters are wrapped in
`fixed()` to record that they were not re-estimated in the joint fit.

# Typical-value structural checks

Before simulating a cohort, confirm the packaged model reproduces the paper's
own arithmetic for a typical subject: 0.8 kg, 33 cm birth length, 25 weeks'
gestational age.

```{r typical-checks}
mod <- readModelDb("Padavia_2024_paracetamol")
mod_typ <- mod |> rxode2::zeroRe()

# Paracetamol molecular weight 151.16 g/mol (a chemical constant; the same
# value van Rongen 2016 prints in its Methods for the identical analyte set).
# Methods 2.4 of Padavia 2024 converts doses to micromoles with the analyte
# molecular weights but does not print them.
MW_APAP <- 151.16
mg_to_umol <- function(mg) mg / MW_APAP * 1000

# A single 20 mg/kg loading dose in a typical 0.8 kg neonate, infused over
# 15 min, then followed to complete washout.
typ_events <- data.frame(
  id = 1L, WT = 0.8, HT_BIRTH = 33, GA = 25,
  time = 0, amt = mg_to_umol(20 * 0.8), evid = 1L, cmt = "central"
) |>
  mutate(rate = amt / 0.25) |>
  bind_rows(
    data.frame(
      id = 1L, WT = 0.8, HT_BIRTH = 33, GA = 25,
      time = c(seq(0, 6, by = 0.02), seq(6.5, 240, by = 0.5)),
      amt = NA_real_, evid = 0L, cmt = "Cc", rate = NA_real_
    )
  ) |>
  arrange(time, desc(evid))

typ <- rxode2::rxSolve(mod_typ, events = typ_events, useLinCmt = FALSE) |>
  as.data.frame()
if (is.null(typ$id)) typ$id <- 1L

# rxSolve returns only observation rows and does not carry an evid column.
pars <- typ |> slice(1)

# `dp` is the number of decimal places the source actually PRINTS for each
# value. Asserting that the model value rounds to the printed value at the
# printed precision is the tightest defensible gate: a flat relative
# tolerance is simultaneously too loose for a 4-decimal entry like CL and too
# tight for a 2-significant-figure one like the 0.81% unchanged fraction,
# whose true value 0.806% is a correct rounding but sits 0.5% away in
# relative terms.
structural <- tibble::tribble(
  ~Quantity,                              ~Published, ~dp, ~Model,
  "CL (L/h)",                             0.0785,     4L,  pars$cl,
  "V1 (L)",                               0.124,      3L,  pars$vc,
  "Q (L/h)",                              3.84,       2L,  pars$q,
  "V2 (L)",                               0.672,      3L,  pars$vp,
  "CL E gluc (L/h)",                      0.0324,     4L,  pars$cle_gluc,
  "CL E sulf (L/h)",                      0.0173,     4L,  pars$cle_sulf,
  "CL E ox (L/h)",                        0.00445,    5L,  pars$cle_cysmer,
  "Sulphate fraction (%)",                89.45,      2L,  100 * pars$fm_sulf,
  "Glucuronide fraction (%)",             6.07,       2L,  100 * pars$fm_gluc,
  "Oxidation fraction (%)",               3.67,       2L,  100 * pars$fm_cysmer,
  "Unchanged fraction (%)",               0.81,       2L,  100 * pars$f_unchanged
) |>
  mutate(
    `Rounds to published` = round(Model, dp) == Published,
    `% diff` = 100 * (Model - Published) / Published
  )

structural |>
  select(-dp) |>
  rename("Published (Table 2 / Results 3.2)" = Published) |>
  knitr::kable(digits = 5, caption = paste(
    "Typical-value parameters and metabolisation fractions for a 0.8 kg,",
    "33 cm, 25-week neonate against the values Padavia 2024 prints.",
    "Each model value is required to round to the printed value at the",
    "precision the source prints it to."
  ))

# Deterministic typical-value quantities from a zeroRe() solve, not cohort
# statistics, so an exact printed-value gate is correct: any mis-transcribed
# Table 2 entry breaks the corresponding row outright.
stopifnot(all(structural$`Rounds to published`))
```

The four metabolisation fractions reproduce the Results 3.2 sentence
"Apparent metabolisation fractions were estimated at 89.45% for sulfate, 6.07%
for glucuronide, 3.67% for oxidation, and 0.81% for unchanged elimination for a
typical preterm neonate of 25 weeks" to the printed precision, which
simultaneously confirms `t1`, `t2`, `t3` and the fraction equations of
Methods 2.4.

A dimensional cross-check on the parent disposition: `V1 + V2` is
`r round(pars$vc + pars$vp, 3)` L, i.e.
`r round((pars$vc + pars$vp) / 0.8, 2)` L/kg, which sits on the ~1 L/kg steady-state
volume paracetamol is known for, and `Vss / CL` gives a mean residence time of
`r round((pars$vc + pars$vp) / pars$cl, 1)` h.

# Replicating Figure 1 -- metabolisation pathways versus gestational age

Figure 1 of Padavia 2024 plots the four elimination pathways against
gestational age over 23-26 weeks. Because the single gestational-age effect sits
on `t1` and `t1` appears in the shared denominator of all four fractions, one
covariate moves the whole split -- the mechanism behind the paper's finding that
"the glucuronide pathway increased and the sulfate pathway decreased" with
maturity.

```{r figure-1}
# The integer gestational ages are unioned in explicitly rather than relied on
# to fall out of the seq(): `seq(23, 26, by = 0.1)` accumulates floating-point
# error and its last element is 25.999999999999996, so a downstream
# `GA %in% c(23, 26)` would match only 23 and the range gate below would
# silently collapse to a single endpoint instead of failing.
ga_grid <- sort(unique(c(as.numeric(23:26), seq(23, 26, by = 0.1))))

ga_events <- data.frame(
  id = seq_along(ga_grid), WT = 0.8, HT_BIRTH = 33, GA = ga_grid,
  time = 0, amt = mg_to_umol(16), evid = 1L, cmt = "central"
) |>
  mutate(rate = amt / 0.25) |>
  bind_rows(
    data.frame(
      id = seq_along(ga_grid), WT = 0.8, HT_BIRTH = 33, GA = ga_grid,
      time = 1, amt = NA_real_, evid = 0L, cmt = "Cc", rate = NA_real_
    )
  ) |>
  arrange(id, time, desc(evid))

ga_sim <- rxode2::rxSolve(mod_typ, events = ga_events, keep = "GA",
                          useLinCmt = FALSE) |>
  as.data.frame()

fractions <- ga_sim |>
  transmute(
    GA,
    Sulphation      = 100 * fm_sulf,
    Glucuronidation = 100 * fm_gluc,
    Oxidation       = 100 * fm_cysmer,
    Unchanged       = 100 * f_unchanged
  ) |>
  pivot_longer(-GA, names_to = "Pathway", values_to = "Percent")

ggplot(fractions, aes(GA, Percent, colour = Pathway)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~Pathway, scales = "free_y") +
  labs(
    x = "Gestational age (weeks)",
    y = "Apparent metabolisation fraction (%)",
    title = "Figure 1 -- paracetamol elimination pathways by gestational age",
    caption = "Replicates Figure 1 of Padavia 2024 (gestational age 23-26 weeks)."
  ) +
  theme(legend.position = "none")
```

Results 3.2 states the range each fraction covers across 23-26 weeks:
sulphation 87.8-91.9%, glucuronidation 3.51-7.8%, oxidation 3.61-3.78% and
unchanged elimination 0.79-0.83%. Both endpoints of all four ranges are printed
values, so this is a gate on the transcription of `t1`, `t2`, `t3` and the
gestational-age exponent, not merely on internal consistency.

```{r figure-1-gate}
endpoints <- fractions |>
  filter(GA %in% c(23, 26)) |>
  pivot_wider(names_from = GA, values_from = Percent, names_prefix = "GA") |>
  mutate(Low = pmin(GA23, GA26), High = pmax(GA23, GA26)) |>
  select(Pathway, Low, High) |>
  arrange(Pathway)

# `dp low` / `dp high` record the decimal places Results 3.2 actually prints
# for each endpoint. Note the asymmetry within a single sentence: the
# glucuronidation range is printed "between 3.51 and 7.8%" and the sulphation
# range "between 87.8 and 91.9%", so the gate must respect one decimal on
# some endpoints and two on others rather than assuming a common precision.
# `dev high` marks the one endpoint the model reproducibly does NOT match --
# the printed oxidation maximum. The chunk after this one shows it is a
# last-digit slip in the source rather than a model defect. It is kept
# visible in the table and excluded from the gate, rather than the gate being
# widened until the disagreement disappears.
published_ranges <- tibble::tribble(
  ~Pathway,           ~`Published low`, ~`dp low`, ~`Published high`, ~`dp high`, ~`dev high`,
  "Glucuronidation",  3.51,             2L,        7.8,               1L,         FALSE,
  "Oxidation",        3.61,             2L,        3.78,              2L,         TRUE,
  "Sulphation",       87.8,             1L,        91.9,              1L,         FALSE,
  "Unchanged",        0.79,             2L,        0.83,              2L,         FALSE
)

range_cmp <- endpoints |>
  left_join(published_ranges, by = "Pathway") |>
  mutate(
    `Low rounds`      = round(Low,  `dp low`)  == `Published low`,
    `High rounds`     = round(High, `dp high`) == `Published high`,
    `Known deviation` = `dev high`
  )

range_cmp |>
  select(-`dp low`, -`dp high`, -`dev high`) |>
  rename("Model low" = Low, "Model high" = High) |>
  knitr::kable(digits = 4, caption = paste(
    "Metabolisation-fraction ranges over gestational age 23-26 weeks,",
    "model versus the ranges printed in Results 3.2 of Padavia 2024.",
    "Each endpoint must round to the printed value at the printed precision;",
    "the one flagged row is a documented rounding slip in the source."
  ))

# Confirm the gate actually had all four pathways and two distinct endpoints
# each to test: a filter matching nothing would leave `all()` vacuously TRUE
# and report a pass on a claim it never evaluated. An earlier draft of this
# vignette hit exactly that -- `seq(23, 26, by = 0.1)` never contains 26
# exactly, so a `GA %in% c(23, 26)` filter silently kept one endpoint.
stopifnot(
  nrow(range_cmp) == 4L,
  !anyNA(range_cmp$`Published low`),
  all(range_cmp$High > range_cmp$Low),
  sum(range_cmp$`Known deviation`) == 1L
)

# Deterministic typical-value quantities from a zeroRe() solve, so the exact
# printed-value gate applies. Seven of the eight endpoints are printed numbers
# the model must reproduce, which makes this a transcription gate on t1, t2,
# t3 and the gestational-age exponent rather than a check of internal
# consistency.
stopifnot(
  all(range_cmp$`Low rounds`),
  all(range_cmp$`High rounds`[!range_cmp$`Known deviation`])
)
```

The one endpoint the model does not reproduce is the printed **oxidation
maximum, 3.78%**, against which the model gives 3.7748% -- a difference in the
last printed digit only. The cause is the rounding of the published parameters
themselves, and it can be shown to be exactly that.

All four fractions at a given gestational age share the single denominator
`D = 1 + t1 + t2 + t3`. Table 2 prints `t2 = 111` to three significant
figures, so the estimate actually used by the authors lies somewhere in
[110.5, 111.5], which moves `D` by up to +/- 0.5. The model has no choice but
to use the printed point value; the paper's own Figure 1 and Results ranges
were computed from the unrounded estimates.

```{r deviation-diagnosis}
t1_23 <- 7.53 * (23 / 25)^6.88
ox_max <- function(t2) 100 * 4.56 / (1 + t1_23 + t2 + 4.56)

t2_probe <- c(110.5, 110.8, 111.0, 111.5)
tibble::tibble(
  `t2` = t2_probe,
  `Oxidation maximum (%)` = ox_max(t2_probe),
  `Rounds to` = round(ox_max(t2_probe), 2)
) |>
  knitr::kable(digits = 4, caption = paste(
    "Oxidation maximum at gestational age 23 weeks as a function of t2,",
    "across the rounding interval of the printed value 111. The printed",
    "3.78% is recovered anywhere below t2 = 110.99."
  ))

# The t2 value at which the oxidation maximum crosses the 3.775 floor of the
# printed 3.78 must lie INSIDE the printed value's rounding interval -- that
# is what makes published-value rounding a sufficient explanation and rules
# out a transcription error on t1 or t3 (either of those would also break the
# glucuronidation and unchanged endpoints, which do reproduce).
t2_cross <- uniroot(function(t2) ox_max(t2) - 3.775, c(105, 115))$root
stopifnot(
  t2_cross > 110.5, t2_cross < 111.5,
  round(ox_max(111.0), 2) == 3.77,
  round(ox_max(110.8), 2) == 3.78
)
cat(sprintf(
  "Oxidation maximum reaches the printed 3.78%% at t2 = %.3f, inside the [110.5, 111.5] rounding interval of the printed 111.\n",
  t2_cross
))
```

So the single non-reproducing endpoint is a consequence of Table 2 reporting
`t2` to three significant figures, not of a transcription error: a
mis-transcribed `t1` or `t3` would move the glucuronide and unchanged
endpoints as well, and both of those reproduce exactly. The model file keeps
the printed point values, which is the correct choice -- inventing a
`t2 = 110.8` to close a last-digit gap would be tuning a parameter to match a
validation target.

# Virtual cohort

The trial data are not public (the source's data-availability statement routes
requests through Inserm), so the cohort below is virtual, with covariate
distributions approximating Table 1: birth weight and birth length drawn from
truncated normals matched to the published medians and interquartile ranges, and
gestational age drawn from the published 2 / 9 / 7 / 12 counts at 23 / 24 / 25 /
26 weeks.

```{r cohort}
# set.seed() seeds R's RNG, which is what draws the covariates below. It does
# NOT seed rxode2's simulation RNG, and rxode2's streams are partitioned per
# solver thread, so the eta draws differ between a 2-core CI runner and a
# many-threaded workstation. Every assertion downstream is therefore written
# to hold for any cohort the model can produce (see pattern 12 of
# references/known-vignette-failure-patterns.md).
set.seed(20241122)

n_per_arm <- 100L

rtnorm <- function(n, mean, sd, lower, upper) {
  x <- rnorm(n, mean, sd)
  pmin(pmax(x, lower), upper)
}

make_cohort <- function(n, load_mgkg, maint_mgkg, arm, id_offset = 0L) {
  subj <- tibble(
    id       = id_offset + seq_len(n),
    arm      = arm,
    # Table 1: birth weight median 800 g, IQR 672.5-857.5, range 470-920.
    # IQR width 185 g implies SD ~ 185 / 1.349.
    WT       = rtnorm(n, 0.800, 0.137, 0.470, 0.920),
    # Table 1: birth length median 32.75 cm, IQR 31.5-34.0, range 28.0-36.5.
    HT_BIRTH = rtnorm(n, 32.75, 1.85, 28.0, 36.5),
    # Table 1 gestational-age counts 2 / 9 / 7 / 12 of 30.
    GA       = sample(c(23, 24, 25, 26), n, replace = TRUE,
                      prob = c(2, 9, 7, 12) / 30)
  )

  # Methods 2.2 / Table 1 footnote: loading dose at t = 0 then maintenance
  # every 6 h, "total 20 doses" over the 5-day course.
  dose_times <- c(0, seq(6, by = 6, length.out = 19L))
  dose_mgkg  <- c(load_mgkg, rep(maint_mgkg, 19L))

  doses <- subj |>
    tidyr::crossing(tibble(time = dose_times, mgkg = dose_mgkg)) |>
    mutate(
      amt  = mg_to_umol(mgkg * WT),
      evid = 1L,
      cmt  = "central",
      # Infusion duration is not reported; 15 min is the standard intravenous
      # paracetamol infusion time. See "Assumptions and deviations".
      rate = amt / 0.25
    ) |>
    select(-mgkg)

  obs <- subj |>
    tidyr::crossing(time = c(seq(0, 12, by = 0.5), seq(13, 144, by = 1))) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "Cc", rate = NA_real_)

  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}

events <- bind_rows(
  make_cohort(n_per_arm, 20, 7.5, "Level 1: 20 then 7.5 mg/kg q6h", id_offset = 0L),
  make_cohort(n_per_arm, 25, 10.0, "Level 2: 25 then 10 mg/kg q6h",
              id_offset = n_per_arm)
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(mod, events = events,
                       keep = c("arm", "WT", "HT_BIRTH", "GA"),
                       useLinCmt = FALSE) |>
  as.data.frame()

stopifnot(nrow(sim) > 0, all(sim$Cc >= 0), all(sim$Cc_sulf >= 0))
```

```{r profiles}
sim |>
  select(id, arm, time, Cc, Cc_gluc, Cc_sulf, Cc_cysmer) |>
  pivot_longer(
    c(Cc, Cc_gluc, Cc_sulf, Cc_cysmer),
    names_to = "Analyte", values_to = "conc"
  ) |>
  mutate(Analyte = recode(Analyte,
    Cc = "Paracetamol", Cc_gluc = "Glucuronide",
    Cc_sulf = "Sulphate", Cc_cysmer = "Cysteine + mercapturate"
  )) |>
  group_by(arm, Analyte, time) |>
  summarise(
    Q05 = quantile(conc, 0.05), Q50 = median(conc),
    Q95 = quantile(conc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(time, Q50, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.20, colour = NA) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~Analyte, scales = "free_y") +
  labs(
    x = "Time (h)", y = "Plasma concentration (umol/L)", colour = NULL, fill = NULL,
    title = "Simulated 5-day profiles, median with 5th-95th percentile band",
    caption = paste(
      "Corresponds to the prediction-corrected VPC panels of supplemental",
      "Figure S4 of Padavia 2024, which are not reproduced numerically here",
      "because the observed data are not public."
    )
  ) +
  theme(legend.position = "bottom")
```

The sulphate metabolite dominates in molar terms and accumulates to a plateau
far above the parent, while the cysteine + mercapturate pool -- fed by only
~3.7% of the parent flux but eliminated slowly
(`CL E ox / V1` gives a half-life of
`r round(log(2) / (0.00445 / 0.124), 1)` h) -- climbs steadily across the 5-day
course. The glucuronide sits between them. This ordering is the pathway split
of Figure 1 expressed as concentrations.

# PKNCA validation

Padavia 2024 reports no NCA table: exposure metrics appear only as boxplots
(Figures 2 and 3) and as per-subject values in supplemental Table S1, neither of
which yields transcribable point estimates. There is therefore nothing to feed
`ncaComparisonTable()`. What *is* available is stronger than a table comparison,
because the model's structure implies exact closed-form identities that
non-compartmental analysis of the simulated profiles must satisfy.

For a linear system given a single intravenous dose `D` (in micromoles) and
followed to complete washout:

* the parent satisfies `CL = D / AUC_parent`, so NCA recovers the clearance;
* each metabolite pool receives `fm_i * D` micromoles and eliminates all of it,
  so `CLE_i * AUC_i = fm_i * D`, i.e. `fm_i = CLE_i * AUC_i / D`.

The second identity is the sharp one: it is only satisfied if the formation
term, the elimination term, the fixed metabolite volume and the mole-for-mole
stoichiometry of the ODE system are all wired correctly. It is computed below
from numerically integrated simulated profiles and compared against each
subject's own `fm_i`, so it tests structure rather than transcription -- the
transcription gates are the two printed-value tables above.

```{r pknca-single-dose}
# A single loading dose per subject, followed for 240 h: the slowest pool
# (cysteine + mercapturate) has a 19.3 h half-life, so 240 h is more than
# twelve half-lives and the terminal phase is well resolved.
# The identities below must hold for every subject individually, so a subset
# of the cohort exercises them just as sharply as the whole of it while
# keeping the fine sampling grid affordable. 25 subjects per arm span the
# drawn range of CL and of all three metabolite clearances.
n_balance <- 25L

sd_doses <- events |>
  filter(evid == 1, time == 0) |>
  group_by(arm) |>
  slice_head(n = n_balance) |>
  ungroup()

# The distribution phase is extremely fast -- Q / V1 is about 31 /h, giving a
# first-phase half-life near one minute -- so most of the parent AUC accrues
# in the first few minutes. A coarse early grid understates AUC0-inf and the
# CL identity below then fails for reasons that have nothing to do with the
# model: at 0.1 h resolution the typical subject already shows a 0.27% error,
# which 0.01 h resolution removes entirely.
sd_grid <- c(seq(0, 2, by = 0.01), seq(2.1, 24, by = 0.1), seq(25, 240, by = 1))

sd_events <- sd_doses |>
  bind_rows(
    sd_doses |>
      select(id, arm, WT, HT_BIRTH, GA) |>
      tidyr::crossing(time = sd_grid) |>
      mutate(amt = NA_real_, evid = 0L, cmt = "Cc", rate = NA_real_)
  ) |>
  arrange(id, time, desc(evid))

sd_sim <- rxode2::rxSolve(mod, events = sd_events,
                          keep = c("arm", "WT"), useLinCmt = FALSE) |>
  as.data.frame()

# Per-subject model quantities needed by the closed-form identities.
sd_pars <- sd_sim |>
  group_by(id, arm) |>
  summarise(
    cl = first(cl), fm_gluc = first(fm_gluc), fm_sulf = first(fm_sulf),
    fm_cysmer = first(fm_cysmer), f_unchanged = first(f_unchanged),
    cle_gluc = first(cle_gluc), cle_sulf = first(cle_sulf),
    cle_cysmer = first(cle_cysmer), .groups = "drop"
  )

dose_df <- sd_doses |>
  select(id, time, amt, arm)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

# One PKNCA block per output (the model has four).
run_nca <- function(analyte_col) {
  # Only `!is.na()` here -- adding a `time > 0` or `Cc > 0` filter at this
  # stage would drop the time-zero row PKNCA needs to anchor AUC0-*.
  d <- sd_sim |>
    filter(!is.na(.data[[analyte_col]])) |>
    transmute(id, time, arm, Cc = .data[[analyte_col]])

  # Trim the far tail to a per-subject numerical noise floor. Followed for
  # 240 h, the parent falls through roughly 34 terminal half-lives and the
  # faster metabolite pools do likewise; those values are solver noise and
  # can go very slightly negative, at which point PKNCA's log-linear
  # lambda-z regression returns NaN and aucinf.obs is lost for that subject.
  # Keeping everything down to 1e-8 of each subject's own peak leaves more
  # than twenty half-lives of genuine signal for the regression.
  d <- d |>
    group_by(id) |>
    filter(time == 0 | Cc > 1e-8 * max(Cc)) |>
    ungroup()

  # Guarantee a time-zero row per subject. Dosing is by infusion into an
  # initially empty system, so the pre-dose value is exactly 0.
  d <- bind_rows(
    d,
    d |> distinct(id, arm) |> mutate(time = 0, Cc = 0)
  ) |>
    distinct(id, arm, time, .keep_all = TRUE) |>
    arrange(id, arm, time)

  conc_obj <- PKNCA::PKNCAconc(d, Cc ~ time | arm + id,
                               concu = "umol/L", timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id, doseu = "umol")
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                        intervals = intervals))
  as.data.frame(res$result) |>
    filter(PPTESTCD %in% c("aucinf.obs", "cmax", "tmax", "half.life")) |>
    select(id, arm, PPTESTCD, PPORRES) |>
    mutate(analyte = analyte_col)
}

nca_all <- bind_rows(lapply(
  c("Cc", "Cc_gluc", "Cc_sulf", "Cc_cysmer"), run_nca
))

auc <- nca_all |>
  filter(PPTESTCD == "aucinf.obs") |>
  select(id, arm, analyte, auc = PPORRES) |>
  pivot_wider(names_from = analyte, values_from = auc)

balance <- auc |>
  left_join(sd_pars, by = c("id", "arm")) |>
  left_join(dose_df |> select(id, D = amt), by = "id") |>
  mutate(
    cl_recovered        = D / Cc,
    fm_gluc_recovered   = cle_gluc   * Cc_gluc   / D,
    fm_sulf_recovered   = cle_sulf   * Cc_sulf   / D,
    fm_cysmer_recovered = cle_cysmer * Cc_cysmer / D,
    err_cl        = 100 * (cl_recovered        - cl)        / cl,
    err_gluc      = 100 * (fm_gluc_recovered   - fm_gluc)   / fm_gluc,
    err_sulf      = 100 * (fm_sulf_recovered   - fm_sulf)   / fm_sulf,
    err_cysmer    = 100 * (fm_cysmer_recovered - fm_cysmer) / fm_cysmer
  )

stopifnot(nrow(balance) == 2L * n_balance, !anyNA(balance$err_cl))

tibble::tibble(
  Identity = c(
    "CL = D / AUC_parent",
    "fm_gluc = CLE_gluc * AUC_gluc / D",
    "fm_sulf = CLE_sulf * AUC_sulf / D",
    "fm_cysmer = CLE_ox * AUC_cysmer / D"
  ),
  `Max |% error| over the subset` = c(
    max(abs(balance$err_cl)), max(abs(balance$err_gluc)),
    max(abs(balance$err_sulf)), max(abs(balance$err_cysmer))
  )
) |>
  knitr::kable(digits = 3, caption = paste(
    "Closed-form mass-balance identities recovered by PKNCA from the",
    "simulated single-dose profiles over a 25-subject-per-arm subset. Both",
    "sides use each subject's own drawn",
    "parameters, so the residual is numerical integration and",
    "lambda-z extrapolation error only."
  ))

# Both sides of each identity use the SAME drawn parameters, so the residual
# is numerical-integration and lambda-z extrapolation error rather than cohort
# spread, and a tight bound is correct -- it is what makes this catch an
# ODE-wiring, volume or unit-conversion mistake, any of which would move a row
# by whole percent or more.
#
# Realised at 1 / 2 / 16 solver threads (identical at all three):
#   parent 0.0017%, glucuronide 0.0034%, sulphate 0.0035%, cysmer 0.4276%.
# The cysmer pool gets a looser bound because it is the slow one: its
# elimination clearance carries the largest IIV in the model (omega 0.926 on
# the log scale), so an unlucky draw has a terminal half-life of order 100 h
# and aucinf.obs must extrapolate further beyond the 240 h window. Do not
# tighten these back to the realised values -- the cohort differs per draw.
stopifnot(
  max(abs(balance$err_cl))     < 0.05,
  max(abs(balance$err_gluc))   < 0.05,
  max(abs(balance$err_sulf))   < 0.05,
  max(abs(balance$err_cysmer)) < 2.0
)

# The four fractions must exhaust the parent clearance.
stopifnot(max(abs(
  balance$fm_gluc + balance$fm_sulf + balance$fm_cysmer +
    balance$f_unchanged - 1
)) < 1e-8)
```

The pathway split recovered from the simulated exposures also reproduces the
paper's headline ordering -- sulphation far ahead of glucuronidation, oxidation
under 4% -- across the whole virtual cohort, not only at the typical value.

```{r nca-summary}
nca_all |>
  filter(PPTESTCD %in% c("cmax", "aucinf.obs", "half.life")) |>
  mutate(
    Analyte = recode(analyte,
      Cc = "Paracetamol", Cc_gluc = "Glucuronide",
      Cc_sulf = "Sulphate", Cc_cysmer = "Cysteine + mercapturate"
    ),
    Parameter = nlmixr2lib::ncaParamLabel(PPTESTCD)
  ) |>
  group_by(arm, Analyte, Parameter) |>
  summarise(
    Median = median(PPORRES, na.rm = TRUE),
    `5th`  = quantile(PPORRES, 0.05, na.rm = TRUE),
    `95th` = quantile(PPORRES, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  rename("Dose level" = arm) |>
  knitr::kable(digits = 2, caption = paste(
    "Simulated single-loading-dose NCA by analyte and dose level.",
    "Cmax in umol/L, AUC0-inf in umol*h/L, half-life in h.",
    "Padavia 2024 publishes no comparable table -- exposures appear only as",
    "boxplots in Figures 2 and 3 and per-subject in supplemental Table S1."
  ))

# Cohort-derived, so assert only what the paper's own structure guarantees for
# any cohort: the sulphate pool must carry the largest molar exposure and the
# oxidative pool the smallest of the three metabolites, because the fractions
# are 89% / 6% / 3.7% and no metabolite clearance is large enough to reverse
# an order-of-magnitude flux difference.
med_auc <- balance |>
  summarise(across(c(Cc_gluc, Cc_sulf, Cc_cysmer), median))
stopifnot(med_auc$Cc_sulf > med_auc$Cc_gluc)
stopifnot(median(balance$fm_sulf) > 10 * median(balance$fm_gluc))
stopifnot(median(balance$fm_cysmer) < 0.06)
```

# Assumptions and deviations

* **Infusion duration is not reported.** Methods 2.2 refers to sampling
  "after the end of the loading dose" and "after the start of infusion", so
  dosing was by infusion, but no duration is given. A 15-minute infusion (the
  standard intravenous paracetamol administration time) is used throughout this
  vignette. Every gate above is AUC- or fraction-based and therefore invariant
  to the input duration for a linear model; only the simulated `Cmax` and
  `Tmax` in the summary table depend on it, and neither is compared against a
  published value.
* **Metabolite volumes are equated to `V1` inside `model()`.** Methods 2.4
  states the metabolite volumes "were fixed to the parent volume" without
  naming which parent volume. `V1` -- the central compartment the metabolites
  are formed from and sampled in -- is the reading taken here; `V2` is a
  peripheral parent compartment with no metabolite counterpart, and equating
  metabolite volumes to a peripheral volume has no mechanistic reading. The
  equality is written in `model()` rather than as three `fixed()` entries in
  `ini()` because it is a structural identifiability constraint of the
  parameterisation, not an estimated value a user should be able to override
  inconsistently.
* **`W` is treated as the time-varying treatment-period weight, held constant
  per subject here.** Methods 2.2 collected body weight daily and Results 3.2
  names the retained covariate "bodyweight" (as distinct from the "birth
  length" retained on `V2`), so the model's `WT` is the daily weight. The
  virtual cohort holds it fixed at the birth value over the 5-day window, which
  neglects the early postnatal weight loss and recovery typical of this
  population. The Table 2 centring value (800 g) coincides exactly with the
  Table 1 birth-weight median, so the typical-value predictions are unaffected.
* **Birth length is registered as a covariate column distinct from `HT`.** The
  dataset carries both a daily body length and a birth length, and only the
  latter was retained, exactly paralleling the existing `WT` / `WT_BIRTH`
  distinction in the covariate register.
* **Covariate centring uses 33 cm, not the Table 1 median of 32.75 cm.** The
  Table 2 footnote states birth length "were centred on median values: ...
  33 cm for birth length", so 33 cm is the value the published equation uses
  and is what the model file encodes.
* **Dose count over the 5-day course.** Methods 2.2 says "20 mg/kg followed by
  7.5 mg/kg four times daily (QID) for 5 days (total 20 doses)". Read literally,
  "total 20 doses" includes the loading dose, which is the reading used here
  (loading dose plus 19 maintenance doses every 6 h); read as "QID for 5 days"
  it would be the loading dose plus 20 maintenance doses. No gate in this
  vignette depends on the choice.
* **`omega` values are standard deviations.** The Table 1 / Table 2 shared
  footnote states "omega inter-subject variability expressed as standard
  deviation", so each printed `omega` is squared to give the variance entered in
  `ini()`. No CV%-to-variance conversion is involved, and no assumption was
  needed -- the scale is stated in print.
* **Paracetamol molecular weight 151.16 g/mol is a chemical constant, not a
  paper value.** Methods 2.4 states that doses and concentrations were converted
  to micromoles using the analyte molecular weights, citing a reference, but
  does not print them. The value used for the mg-to-micromole conversion in
  this vignette is the standard molecular weight of paracetamol, and is the same
  value the sibling model `vanRongen_2016_acetaminophen` prints from its own
  Methods for the identical analyte set. It affects only the dose amounts in
  this vignette, not any `ini()` parameter.
* **One printed value the model does not reproduce: the oxidation-fraction
  maximum.** Results 3.2 gives the oxidation range across 23-26 weeks as
  "between 3.61 and 3.78%"; the model reproduces the 3.61% lower endpoint
  exactly and gives 3.7748% against the printed 3.78% upper endpoint. This is
  a consequence of Table 2 printing `t2` to three significant figures (`111`):
  any value below `t2 = 110.99`, comfortably inside that rounding interval,
  reproduces 3.78%, and the diagnosis chunk above demonstrates this. The model
  file keeps the printed point estimates rather than substituting a `t2` chosen
  to close the gap, which would be tuning a parameter to match a validation
  target. The remaining seven of the eight printed range endpoints reproduce at
  the printed precision, as do all eleven typical-value quantities.
* **No published NCA comparison table.** As described above, the source reports
  exposures only graphically and in a supplement not available on disk, so
  `nlmixr2lib::ncaComparisonTable()` is not used. The closed-form mass-balance
  identities and the two printed-value tables (typical-value parameters and
  fractions; fraction ranges across gestational age) are the validation
  instead. Supplemental Figure 1 (the structural-model diagram), Figures S2-S3
  (diagnostics), Figure S4 (prediction-corrected VPCs) and Table S1
  (per-subject exposures) were not obtainable; the five-compartment structure
  those figures illustrate is fully determined by the Methods 2.4 equations and
  the Results 3.2 description, and every `ini()` value traces to the printed
  Table 2.
* **The clinical-outcome analyses are not part of the model.** Sections 3.4 and
  the associated Figures 2 and 3 report Spearman correlations and Wilcoxon tests
  between exposure and liver biomarkers or ductus closure, all of which were
  null. These are statistical tests on derived exposures rather than a fitted
  exposure-response model, so there is no PD sub-model to extract.
* **Extrapolation outside 23-26 weeks is not supported.** The gestational-age
  exponents are steep (6.88 on `t1`, 6.26 on `CL E sulf`) and were estimated
  over a three-week window. The paper itself notes only that "a reversal between
  those two metabolisation fractions would occur by extrapolating to older
  ages"; treating the power functions as valid at term gestation is not
  warranted by these data.
