---
title: "Gentamicin (Paioni 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gentamicin (Paioni 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r readModelDb("Paioni_2021_gentamicin")()$reference`
* Article: <https://doi.org/10.3390/pharmaceutics13101596> (open access)
* The online supplement holds figures, the research plan and the start-value
  ranges of the multistart fit (Table S1); every value used below comes from the
  main article text, equations and tables. No correction notice was found for
  this article as of 2026-09-29.

# Population

Paioni 2021 is a prospective observational study at University Children's
Hospital Zurich (1 October 2017 to 30 April 2019). It enrolled 109 children
aged 0.9 days to 14.5 years (median 29.1 days) receiving gentamicin for at least
48 h; 67 were male and 42 female. Median body weight was 4.23 kg (range
2.5-35, mean 5.42), median gestational age 39.3 weeks (29-42; 11 born before 37
weeks). Serum creatinine was below the 27 umol/L reporting limit in 73 of 107
patients (range < 27-72 umol/L) and median serum urea was 3.20 mmol/L (< 1.8 to
17.3) (Table 1). Indications were suspected (43), suspected superimposed (14)
and proven (51) bacterial infection, plus one surgical prophylaxis (Table 2).
Patients with cystic fibrosis or on any dialysis / haemofiltration were
excluded.

Gentamicin was given once daily per hospital practice (5 mg/kg at age < 7 days,
7.5 mg/kg at >= 7 days) as a 2-min or 30-min infusion. Up to three samples were
taken per dosing interval (trough, about 30 min and about 4 h after the end of
the second or third infusion): 310 concentrations, 71 below the 0.3 mg/L LOQ.

The same information is available programmatically via
`readModelDb("Paioni_2021_gentamicin")()$population`.

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Two-compartment model, zero-order infusion, linear distribution and elimination | n/a | Section 2.3, Equations (1)-(5); Section 3.2 |
| `lambda_z = CL / (V1 + V2')` | n/a | Section 2.3 text following Equation (1) |
| `f1 = (1/V1) * (k21 - lambda1)/(lambda_z - lambda1) * CL/lambda1` | n/a | Equation (2) |
| `k21 = lambda1 * lambda_z * V1 / CL` | n/a | Equation (3) |
| Covariate equations (power of `cov / ref`) | n/a | Table 3 equation rows |
| `lcl` | 2.267 (ln L/d) | Table 3, q1 |
| `lvparea` (V2') | -0.174 (ln L) | Table 3, q2 |
| `lalpha` (lambda1) | 3.644 (ln 1/d) | Table 3, q3 |
| `lvc` (V1) | 0.204 (ln L) | Table 3, q4 |
| `e_wt_cl` | 1.219 | Table 3, q5 (reference ln(4 kg)) |
| `e_creat_cl` | -0.964 | Table 3, q6 (reference ln(27 uM)) |
| `e_bun_cl` | -0.168 | Table 3, q7 (reference ln(3 mM)) |
| `e_wt_vparea` | 0.974 | Table 3, q8 (reference ln(4 kg)) |
| `e_wt_vc` | 0.688 | Table 3, q9 (reference ln(4 kg)) |
| `etalcl` | 0.107 (variance) | Table 3, variance of random effects ln(CL) |
| `etalvparea` | 0.291 (variance) | Table 3, variance of random effects ln(V2') |
| `etalvc` | 0.0391 (variance) | Table 3, variance of random effects ln(V1) |
| No IIV on lambda1 | n/a | Sections 3.2 and 3.3 |
| `expSd` | 0.102 | Table 3, residual error; "exponential" error model, Section 2.3 |

## Parameterisation

The authors fit the closed-form bi-exponential solution (Equation 1) with four
parameters: CL, the central volume V1, the fast disposition rate constant
lambda1, and an *apparent* peripheral volume V2' defined so that the terminal
rate constant is lambda_z = CL / (V1 + V2'). V1 + V2' is therefore the
terminal-phase volume (V_z, "V_area"), not the steady-state volume, and V2' is
not the peripheral volume of the micro-constant parameterisation. The packaged
model keeps the authors' four parameters (so every `ini()` value appears in
Table 3) and derives the micro-constants from the two-compartment identities

* kel = CL / V1
* k21 = lambda1 * lambda_z / kel (Equation 3)
* k12 = lambda1 + lambda_z - kel - k21

The derived peripheral volume `vp = V1 * k12 / k21` and inter-compartmental
clearance `q = k12 * V1` are carried as model outputs. For the typical 4-kg
patient, V2' = 0.840 L while the micro-constant peripheral volume is 0.667 L.

The chunk below checks that the ODE model reproduces the paper's closed form
(Equations 1, 2, 4 and 5) for a typical patient given three 30-min infusions.
Both sides use the same parameters, so the agreement is limited only by the ODE
solver tolerance.

```{r closed-form}
mod <- readModelDb("Paioni_2021_gentamicin")
mod_typ <- rxode2::zeroRe(mod)

typ_cov <- data.frame(WT = 4, CREAT = 27, BUN = 3)
dose_mg <- 7.5 * typ_cov$WT
tinf <- 0.5
obs_t <- sort(unique(c(seq(0, 72, by = 0.25), 48 + tinf + 0.5)))

ev_typ <- dplyr::bind_rows(
  data.frame(time = c(0, 24, 48), evid = 1, amt = dose_mg, dur = tinf, cmt = "central"),
  data.frame(time = obs_t, evid = 0, amt = 0, dur = 0, cmt = "central")
) |>
  dplyr::mutate(id = 1L, WT = typ_cov$WT, CREAT = typ_cov$CREAT, BUN = typ_cov$BUN) |>
  dplyr::arrange(time, dplyr::desc(evid))

sim_typ <- rxode2::rxSolve(mod_typ, events = ev_typ, returnType = "data.frame")

# Paper Equation (1), time in hours. Parameters from Table 3 at the reference
# covariates, converted from per-day to per-hour.
cl <- exp(2.267) / 24
v1 <- exp(0.204)
v2a <- exp(-0.174)
l1 <- exp(3.644) / 24
lz <- cl / (v1 + v2a)
k21 <- l1 * lz * v1 / cl
f1 <- (1 / v1) * (k21 - l1) / (lz - l1) * cl / l1
r0 <- dose_mg / tinf
closed_form <- function(t) {
  out <- 0
  for (t0 in c(0, 24, 48)) {
    tin <- (t - t0) * ((t - t0) <= tinf) * ((t - t0) >= 0)
    tel <- (t - t0 - tinf) * ((t - t0 - tinf) >= 0)
    after <- (tinf + t0) < t
    term <- function(lam) {
      (1 - exp(-lam * tin)) + (1 - exp(-lam * tinf)) * exp(-lam * tel) * after
    }
    out <- out + r0 / cl * (f1 * term(l1) + (1 - f1) * term(lz))
  }
  out
}

chk <- sim_typ |>
  dplyr::filter(time > 0) |>
  dplyr::mutate(Cc_eq1 = closed_form(time), rel_diff = (Cc - Cc_eq1) / Cc_eq1)

c(f1 = f1, max_abs_rel_diff = max(abs(chk$rel_diff)))
stopifnot(max(abs(chk$rel_diff)) < 1e-3)
```

# Virtual cohort

The individual data are not public (they are held on the SPHN Leonhard Med
platform). The simulations below draw a virtual cohort whose covariate
distributions approximate Table 1:

* body weight: log-normal with location log(3.0 kg) and log-SD 0.7, truncated
  to the observed 2.5-35 kg range (after truncation this gives a median of
  about 4.3 kg and a mean of about 5.3 kg, against 4.23 and 5.42 kg in
  Table 1);
* serum creatinine: 68% (73/107) at the 27 umol/L reporting limit, the rest
  log-uniform between 27 and 72 umol/L;
* serum urea: log-normal, median 3.2 mmol/L and log-SD 0.45, truncated to
  1.8-17.3 mmol/L.

Covariates are drawn independently; the paper does not report their joint
distribution.

```{r cohort}
set.seed(20211001)
n_per_arm <- 200L

draw_trunc_lnorm <- function(n, meanlog, sdlog, lo, hi) {
  out <- numeric(0)
  while (length(out) < n) {
    x <- rlnorm(n, meanlog, sdlog)
    out <- c(out, x[x >= lo & x <= hi])
  }
  out[seq_len(n)]
}

make_covariates <- function(n) {
  creat_censored <- runif(n) < 73 / 107
  data.frame(
    WT = draw_trunc_lnorm(n, log(3.0), 0.7, 2.5, 35),
    CREAT = ifelse(creat_censored, 27, exp(runif(n, log(27), log(72)))),
    BUN = draw_trunc_lnorm(n, log(3.2), 0.45, 1.8, 17.3)
  )
}

covs <- make_covariates(n_per_arm)
summary(covs)
```

## Typical-value PK parameters across the cohort (Table 4)

Table 4 of the paper summarises the population-level (covariate-only) PK
parameters of the 109 study patients. The same quantities computed for the
virtual cohort:

```{r table4}
pk_typ <- covs |>
  dplyr::mutate(
    cl_Lh = exp(2.267 + 1.219 * log(WT / 4) - 0.964 * log(CREAT / 27) - 0.168 * log(BUN / 3)) / 24,
    v1 = exp(0.204 + 0.688 * log(WT / 4)),
    v2a = exp(-0.174 + 0.974 * log(WT / 4)),
    vz = v1 + v2a,
    CL_mLmin = cl_Lh * 1000 / 60,
    t_half = log(2) * vz / cl_Lh
  )

tab4 <- tibble::tribble(
  ~parameter, ~published_median, ~published_range,
  "CL (mL/min)", 6.88, "2.76-73.2",
  "Vz = V1 + V2' (L)", 2.16, "1.42-12.4",
  "V1 (L)", 1.27, "0.887-5.46",
  "t1/2 (h)", 3.53, "1.72-8.38"
) |>
  dplyr::mutate(
    simulated_median = c(
      median(pk_typ$CL_mLmin), median(pk_typ$vz),
      median(pk_typ$v1), median(pk_typ$t_half)
    ),
    simulated_range = c(
      paste(signif(range(pk_typ$CL_mLmin), 3), collapse = "-"),
      paste(signif(range(pk_typ$vz), 3), collapse = "-"),
      paste(signif(range(pk_typ$v1), 3), collapse = "-"),
      paste(signif(range(pk_typ$t_half), 3), collapse = "-")
    ),
    pct_diff = 100 * (simulated_median - published_median) / published_median
  )

tab4 |>
  dplyr::rename(
    "Parameter" = parameter,
    "Published median" = published_median,
    "Published range" = published_range,
    "Virtual-cohort median" = simulated_median,
    "Virtual-cohort range" = simulated_range,
    "% difference of medians" = pct_diff
  ) |>
  knitr::kable(digits = 2, caption = "Population-level PK parameters: Table 4 of Paioni 2021 vs the virtual cohort.")

# Structural check on the medians: a mis-transcribed intercept or exponent
# moves these by tens of percent. Weight is the dominant covariate and the
# virtual cohort matches the Table 1 weight median only approximately.
stopifnot(all(abs(tab4$pct_diff) < 10))
```

## Scale of the Table 3 random effects

Table 3 labels the inter-individual terms "Variance of the random effects",
the quantity saemix reports. Table 4 independently confirms that reading: it
prints each PK parameter at the population level (covariates only) and at the
individual level (covariates plus the fitted random effects), and the gap
between the two extremes fixes a minimum size for the random effect of at
least one patient. Population-level V1 per 70 kg is largest at the lightest
weight (2.5 kg), where the model gives the Table 4 value of 24.8 L/70 kg, yet
one patient's individual V1 reaches 33.4 L/70 kg. Likewise the lowest
individual CL is 19.7 mL/min per 70 kg against a population-level minimum of
42.4.

```{r omega-scale}
v1_70_max <- exp(0.204) * (2.5 / 4)^0.688 * 70 / 2.5
eta_v1_min <- log(33.4 / 24.8)
eta_cl_min <- log(19.7 / 42.4)
omega_check <- tibble::tibble(
  parameter = c("V1", "CL"),
  printed_value = c(0.0391, 0.107),
  min_abs_eta = abs(c(eta_v1_min, eta_cl_min))
) |>
  dplyr::mutate(
    z_if_variance = min_abs_eta / sqrt(printed_value),
    z_if_sd = min_abs_eta / printed_value
  )
omega_check |>
  dplyr::rename(
    "Parameter" = parameter,
    "Table 3 value" = printed_value,
    "Smallest |eta| implied by Table 4" = min_abs_eta,
    "|eta| / SD, value read as variance" = z_if_variance,
    "|eta| / SD, value read as SD" = z_if_sd
  ) |>
  knitr::kable(digits = 2, caption = "Random-effect size implied by the Table 4 individual-level extremes.")

# Deterministic (no cohort): the population-level bound reproduces Table 4, and
# only the variance reading keeps the implied random effects plausible.
stopifnot(
  abs(v1_70_max / 24.8 - 1) < 0.01,
  all(omega_check$z_if_variance < 3),
  all(omega_check$z_if_sd > 5)
)
```

Read as SDs, the printed values would require random effects of 7 or more
standard deviations, so the model encodes them as variances.

# Simulation of the typical once-daily regimens (Figure 7)

Section 3.7 simulates three 24-hourly doses of 3, 4, 5 and 7.5 mg/kg in the
study population with the random effects of Table 3 and no residual error, and
reports the 95% interval of the concentration 30 min after the end of the third
infusion, C(30 min'). The virtual cohort above is dosed with each regimen as a
30-min infusion, and with 7.5 mg/kg as a 2-min infusion.

```{r simulate}
regimens <- tibble::tribble(
  ~regimen, ~mgkg, ~tinf,
  "3 mg/kg, 30 min", 3, 0.5,
  "4 mg/kg, 30 min", 4, 0.5,
  "5 mg/kg, 30 min", 5, 0.5,
  "7.5 mg/kg, 30 min", 7.5, 0.5,
  "7.5 mg/kg, 2 min", 7.5, 2 / 60
)

make_arm <- function(i, covs) {
  r <- regimens[i, ]
  n <- nrow(covs)
  ids <- (i - 1L) * n + seq_len(n)
  obs_t <- sort(unique(c(seq(0, 72, by = 0.25), 48 + r$tinf + 0.5, 72)))
  subj <- covs |> dplyr::mutate(id = ids, regimen = r$regimen)
  doses <- subj |>
    tidyr::crossing(time = c(0, 24, 48)) |>
    dplyr::mutate(evid = 1L, amt = r$mgkg * WT, dur = r$tinf, cmt = "central")
  obs <- subj |>
    tidyr::crossing(time = obs_t) |>
    dplyr::mutate(evid = 0L, amt = 0, dur = 0, cmt = "central")
  dplyr::bind_rows(doses, obs)
}

events <- dplyr::bind_rows(lapply(seq_len(nrow(regimens)), make_arm, covs = covs)) |>
  dplyr::arrange(id, time, dplyr::desc(evid))
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))

rxode2::rxSetSeed(20211001)
sim <- rxode2::rxSolve(mod, events = events, keep = c("regimen", "WT"), returnType = "data.frame") |>
  dplyr::mutate(regimen = factor(regimen, levels = regimens$regimen))

# Sensitivity run: the Table 3 values used as standard deviations (so each
# variance is the square of the printed value). See "C(30 min') after the
# third dose" below for why.
mod_sd <- mod |>
  rxode2::ini(etalcl ~ 0.011449, etalvparea ~ 0.084681, etalvc ~ 0.00152881)
rxode2::rxSetSeed(20211001)
sim_sd <- rxode2::rxSolve(mod_sd, events = events, keep = c("regimen", "WT"), returnType = "data.frame") |>
  dplyr::mutate(regimen = factor(regimen, levels = regimens$regimen))
```

```{r figure-7, fig.height = 6}
# Replicates Figure 7 (a, d, g, j) of Paioni 2021: 95% interval and median of
# the simulated concentration-time curves for 30-min infusions.
sim |>
  dplyr::filter(grepl("30 min", regimen)) |>
  dplyr::group_by(regimen, time) |>
  dplyr::summarise(
    lo = quantile(Cc, 0.025), med = median(Cc), hi = quantile(Cc, 0.975),
    .groups = "drop"
  ) |>
  ggplot(aes(time, med)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey70") +
  geom_line(colour = "red") +
  geom_hline(yintercept = c(1, 12), colour = "blue", linetype = "dashed") +
  facet_wrap(~regimen) +
  labs(
    x = "Time after the start of the first infusion (h)",
    y = "Gentamicin plasma concentration (mg/L)",
    caption = "Replicates Figure 7 (a, d, g, j) of Paioni 2021. Grey, 95% interval; red, median."
  )
```

## C(30 min') after the third dose

```{r c30-table}
summarise_c30 <- function(s, label) {
  s |>
    dplyr::mutate(t30 = 48 + regimens$tinf[match(as.character(regimen), regimens$regimen)] + 0.5) |>
    dplyr::filter(abs(time - t30) < 1e-6) |>
    dplyr::group_by(regimen) |>
    dplyr::summarise(
      sim_lo = quantile(Cc, 0.025), sim_med = median(Cc), sim_hi = quantile(Cc, 0.975),
      .groups = "drop"
    ) |>
    dplyr::mutate(iiv = label)
}
c30 <- dplyr::bind_rows(
  summarise_c30(sim, "Table 3 as variances (packaged model)"),
  summarise_c30(sim_sd, "Table 3 as SDs (sensitivity)")
)

pub_c30 <- tibble::tribble(
  ~regimen, ~pub_lo, ~pub_hi,
  "3 mg/kg, 30 min", 5.2, 8.4,
  "4 mg/kg, 30 min", 6.9, 11.3,
  "5 mg/kg, 30 min", NA, NA,
  "7.5 mg/kg, 30 min", 13.0, 21.2,
  "7.5 mg/kg, 2 min", 14.6, 25.0
)

c30_cmp <- c30 |>
  dplyr::mutate(regimen = as.character(regimen)) |>
  dplyr::left_join(pub_c30, by = "regimen") |>
  dplyr::mutate(
    pub_mid = sqrt(pub_lo * pub_hi),
    pct_diff_mid = 100 * (sim_med - pub_mid) / pub_mid
  )

c30_cmp |>
  dplyr::select(iiv, regimen, pub_lo, pub_hi, sim_lo, sim_med, sim_hi, pct_diff_mid) |>
  dplyr::rename(
    "IIV" = iiv,
    "Regimen" = regimen,
    "Published 2.5%" = pub_lo,
    "Published 97.5%" = pub_hi,
    "Simulated 2.5%" = sim_lo,
    "Simulated median" = sim_med,
    "Simulated 97.5%" = sim_hi,
    "% diff of median vs published interval midpoint" = pct_diff_mid
  ) |>
  knitr::kable(digits = 1, caption = "C(30 min') after the third dose (mg/L): Section 3.7 of Paioni 2021 vs simulation.")

# The paper does not print the median, so the centre is compared with the
# geometric midpoint of the published 95% interval. A mis-transcribed volume or
# clearance moves this by tens of percent.
stopifnot(all(abs(c30_cmp$pct_diff_mid) < 12, na.rm = TRUE))
```

The centre of every published interval is reproduced. The published intervals
are, however, markedly narrower than those of the packaged model (for example
5.2-8.4 mg/L at 3 mg/kg), and they are matched closely when the Table 3 values
are used as standard deviations instead of variances. The Table 4 check above
shows the fitted random effects are variances, so the paper's own Section 3.7
simulations appear to have drawn the random effects with the printed variance
used as the standard deviation (Section 2.7 describes drawing from N(0, s^2)
"with s equal the standard deviation of the respective random effect"). The
packaged model keeps the fitted variances; its C(30 min') and C(24 h')
intervals are therefore wider than those printed in the paper. The paper also
reports the 99% interval of C(30 min') at 5 mg/kg (8.2-16.9 mg/L) and at
7.5 mg/kg (12.4-25.4 mg/L), both for 30-min infusions (Discussion).

```{r c24}
# Section 3.7 / Discussion: for doses up to 4 mg/kg, >= 95% of simulated
# C(24 h') (24 h after the start of the third infusion) were below 1 mg/L;
# abstract: at 7.5 mg/kg 95% were < 1.8 mg/L.
summarise_c24 <- function(s, label) {
  s |>
    dplyr::filter(time == 72) |>
    dplyr::group_by(regimen) |>
    dplyr::summarise(
      q95 = quantile(Cc, 0.95),
      pct_below_1 = 100 * mean(Cc < 1),
      .groups = "drop"
    ) |>
    dplyr::mutate(iiv = label, .before = 1)
}
c24 <- dplyr::bind_rows(
  summarise_c24(sim, "Table 3 as variances (packaged model)"),
  summarise_c24(sim_sd, "Table 3 as SDs (sensitivity)")
)
c24 |>
  dplyr::rename(
    "IIV" = iiv,
    "Regimen" = regimen,
    "95th percentile (mg/L)" = q95,
    "% below 1 mg/L" = pct_below_1
  ) |>
  knitr::kable(digits = 2, caption = "Simulated C(24 h') after the third dose.")
```

# PKNCA validation

The paper reports no NCA table. Its Table 4 gives the median terminal
half-life of the study patients at the population level (covariates only, no
random effects), 3.53 h. PKNCA is therefore run over the first dosing interval
(0-24 h) of each 30-min regimen on a typical-value solve of the virtual cohort
(random effects set to zero), which is the same quantity.

```{r pknca}
ev_30 <- events |> dplyr::filter(grepl("30 min", regimen))
# rxode2 warns that a multi-subject solve has no omega; that is the intent here.
sim_pop <- suppressWarnings(
  rxode2::rxSolve(mod_typ, events = ev_30, keep = c("regimen"), returnType = "data.frame")
)

sim_nca <- sim_pop |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(regimen = as.character(regimen)) |>
  dplyr::select(id, time, Cc, regimen)

sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, regimen) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, regimen, time, .keep_all = TRUE) |>
  dplyr::arrange(id, regimen, time)

dose_df <- ev_30 |>
  dplyr::filter(evid == 1, time == 0) |>
  dplyr::select(id, time, amt, dur, regimen)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | regimen + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | regimen + id, route = "intravascular", duration = "dur")
intervals <- data.frame(start = 0, end = 24, cmax = TRUE, tmax = TRUE, auclast = TRUE, half.life = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

published <- tibble::tibble(
  regimen = unique(sim_nca$regimen),
  half.life = 3.53
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "regimen",
  units = c(cmax = "mg/L", tmax = "h", auclast = "h*mg/L", half.life = "h"),
  tolerance_pct = 20
)
cmp |>
  knitr::kable(caption = "PKNCA over the first dosing interval vs the Table 4 median half-life. * differs from reference by >20%.")
```

The model is linear, so the half-life does not depend on the dose; the four
rows differ only through the dose-proportional grid. The residual difference
from 3.53 h reflects the virtual cohort's covariate mix (Table 4 median weight
and creatinine are those of the study patients).

# Assumptions and deviations

- **Parameter names.** The paper's apparent peripheral volume V2' is carried as
  `lvparea` / `vparea` and the distribution-phase rate constant lambda1 as
  `lalpha` / `alpha`; the micro-constants and the micro-constant peripheral
  volume `vp` are derived (see "Parameterisation").
- **Units.** Table 3 reports CL in L/day and lambda1 in 1/day; the model
  converts both to hours.
- **Residual error.** saemix's "exponential" error model is log(y) = log(f) +
  a * eps; the Table 3 value 0.102 is taken as the log-scale SD `a`, the
  quantity saemix reports, and encoded as `lnorm(expSd)`.
- **Censored covariates.** Serum creatinine was below the 27 umol/L reporting
  limit in 73 of 107 patients and the reference value is 27 umol/L; the paper
  does not say how the censored values were coded. Entering them at 27 umol/L
  is assumed (and is what the virtual cohort does). Supplying a creatinine
  below 27 umol/L extrapolates the power effect beyond the data. The coding of
  the eight urea values below 1.8 mmol/L is also not stated.
- **Missing covariates** were entered at the reference value in the fit
  (Section 2.3); do the same when simulating a patient with a missing value.
- **Scale of the random effects.** Table 3 values are variances (Table 3
  heading; confirmed by the Table 4 individual-level extremes). The paper's
  simulated C(30 min') intervals (Section 3.7, abstract, Discussion) are
  reproduced in their centre but are narrower than the packaged model's; they
  match a simulation that uses the printed values as SDs, which suggests the
  published simulations used that scale. The model is not adjusted to match
  them.
- **Structural constraint.** The fit rejected parameter sets with CL >
  lambda1 * V1 (Section 2.3). For such a set the derived k12 is negative; the
  ODE still reproduces the paper's bi-exponential Equation (1) but the
  peripheral amount is no longer physical. This only occurs far outside the
  covariate range of the study (for example a heavy child with a low
  creatinine and a high CL random effect).
- **Virtual cohort.** Covariates were drawn independently from distributions
  matched to the Table 1 summary statistics; the study's own 109 patients (whom
  the paper's simulations used, 100 replicates each) are not available.
- **BLQ handling** in the fit (Section 2.4) affects estimation only and has no
  counterpart in the simulation model. The bracketed Table 3 values from the
  LOQ/2 sensitivity analysis are not used.
