---
title: "Moxifloxacin (Palmer 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Moxifloxacin (Palmer 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Palmer_2025_moxifloxacin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Palmer_2025_moxifloxacin"))$description`
* Article: <https://doi.org/10.1002/bcp.70005>
* Supplement (Supplementary Material 1-9, including the final NONMEM control
  stream in Supplementary Material 9):
  <https://www.ebi.ac.uk/europepmc/webservices/rest/PMC12122143/supplementaryFiles>

# Population

CATALYST was an open-label multisite trial of child-friendly generic
formulations of moxifloxacin and clofazimine (Pan African Clinical Trials
Registry 202012756409365). Thirty-six children being treated for
rifampicin-resistant tuberculosis were enrolled across three sites: South Africa
(n = 20), the Philippines (n = 10) and India (n = 6). Median age was 4.8 years
(range 0.4-15) and median weight 15.6 kg (range 6.9-42.1); enrolment was
stratified into two parallel weight cohorts so that 16 children weighed under
15 kg, of whom 6 weighed under 10 kg. Fourteen (39%) were male. Median
height-for-age z-score was -1.1 (range -4.1 to 0.46) and median weight-for-age
z-score -1.6 (range -4.7 to 1.2). Tuberculosis was microbiologically confirmed
in 17 (47.2%). Only one child (3%) was living with HIV (Table 1).

Each child was sampled intensively twice. At visit PK1 they were taking the
standard non-dispersible 400 mg adult tablet, crushed and suspended in water;
at visit PK2, one to fourteen days later, they took their first dose of the
dispersible 100 mg scored paediatric tablet, dispersed in water. Samples were
drawn predose and at 1, 2, 4, 8 and 24 h postdose on each visit day, with
children fasted for at least 4 h before the predose draw and then given
breakfast before dosing. Doses followed the WHO weight-band recommendations
(Methods 2.1.3). Moxifloxacin was assayed by validated LC-MS/MS with an LLOQ of
0.0628 ug/mL.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Palmer_2025_moxifloxacin")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Palmer_2025_moxifloxacin.R`. The table
below collects them in one place for review. Every number below was read from
the Palmer 2025 main article; the model *structure* was confirmed against the
final NONMEM control stream printed in Supplementary Material 9.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` | log(6.90 L/h) | Table 2, `CL/F (L/h)` |
| `lvc` | log(61.1 L) | Table 2, `Vc/F (L)` |
| `lq` | log(0.860 L/h) | Table 2, `Q/F (L/h)` |
| `lvp` | log(44.5 L) | Table 2, `Vp/F (L)` |
| `lmat` | log(1.01 h) | Table 2, `MAT (h)` |
| `lfdepot` | fixed(log(1)) | Table 2, `F` = 1 (fixed); footnote a |
| `e_wt_cl` | fixed(0.75) | Table 2 footnote a (exponent for CL and Q) |
| `e_wt_vc` | fixed(1.0) | Table 2 footnote a (exponent for Vc and Vp) |
| `e_selfadmin_iovscale` | 1.70 | Table 2, `Fold change of unobserved dosing event on IOV of F and MAT` |
| `etalcl` | 0.136^2 = 0.018496 | Table 2, `IIV on CL (%CV)` = 13.6; footnote c defines %CV as `sqrt(variance) * 100%` |
| `etaiov_fdepot_1..4` | 0.308^2 = 0.094864 | Table 2, `IOV on F (%CV)` = 30.8; footnote c |
| `etaiov_mat_1..4` | 0.422^2 = 0.178084 | Table 2, `IOV on MAT (%CV)` = 42.2; footnote c |
| `propSd` | 0.136 | Table 2, `Proportional error (%)` = 13.6; footnote c |
| `addSd` | 0.0435 mg/L | Table 2, `Additive error (mg/L)`; footnote d, "reported in the standard deviation scale" |
| Allometry `(WT/16)^theta` on CL, Q, Vc, Vp | n/a | Table 2 footnote a |
| `ktr = 4 / mat`; `depot -> transit1 -> transit2 -> transit3 -> central` | n/a | Supplementary Material 9 `$PK`: `KTR = 4/MAT`, `K14 = K45 = K56 = K62 = KTR`; `$MODEL NCOMP=6 COMP=(DOS,DEFDOSE) COMP=(MOXC,DEFOBS) COMP=(MOXP) COMP=(TRANSI) COMP=(TRANSI2) COMP=(TRANSI3)`; Results paragraph 2 ("absorption via 3 transit compartments and estimating mean absorption time") |
| `k12 = q/vc`, `k21 = q/vp`, `kel = cl/vc` | n/a | Supplementary Material 9 `$PK`: `K23 = Q/V2`, `K32 = Q/V3`, `K20 = CL/V2` |
| `f(depot) = exp(lfdepot + iov_fdepot)` | n/a | Supplementary Material 9 `$PK`: `F1 = 1 * EXP(IOVF * UNOBS_EFF)` |
| `mat = exp(lmat + iov_mat)`; `iov_*_scale = 1.70^SELFADMIN` | n/a | Supplementary Material 9 `$PK`: `UNOBS_EFF = 1 + THETA(10)*UNOBS`, `MAT = TVMAT * EXP(IOVMAT * UNOBS_EFF)` |
| `Cc ~ add(addSd) + prop(propSd)` | n/a | Supplementary Material 9 `$ERROR`: `Y = IPRED + IPRED*EPS(1) + EPS(2)` |

The final model carries **no formulation term**: the dispersible-versus-standard
ratios on bioavailability (1.05, 90% CI 0.95-1.15) and on MAT (1.06, 90% CI
0.90-1.26) were not significant, and Table 2 has no formulation row.

```{r model}
mod <- readModelDb("Palmer_2025_moxifloxacin")
th <- rxode2::rxode(mod)$theta
mod_typical <- rxode2::zeroRe(mod)

tau <- 24      # once-daily dosing interval, h
n_dose <- 12   # doses simulated to reach steady state
t_end <- tau * n_dose
```

# Structural checks

Three relationships in this paper are exact arithmetic on the packaged
parameters rather than statements about a simulated cohort, so they can be
checked to numerical precision.

## Table 4 dose ratios

The paper states that, relative to the WHO recommendations, the proposed doses
were "adjusted upwards by 50-56% in 5-7 kg for both formulations and 33% in
7-10 kg for standard formulation; doses in children >10 kg remained the same as
WHO recommended". That is pure arithmetic on Table 4 and must reproduce exactly.

```{r table4}
# Palmer 2025 Table 4, daily dose in mg by WHO weight band.
table4 <- tibble::tibble(
  band = factor(rep(c("5-<7", "7-<10", "10-<16", "16-<24", ">=24"), 2),
                levels = c("5-<7", "7-<10", "10-<16", "16-<24", ">=24")),
  formulation = rep(c("Standard 400 mg", "Dispersible 100 mg"), each = 5),
  # Band midpoints; the top band is simulated over 24-31 kg because Results 3.2
  # notes the >=25 kg recommendation equals the 400 mg adult dose.
  wt_lo = rep(c(5, 7, 10, 16, 24), 2),
  wt_hi = rep(c(7, 10, 16, 24, 31), 2),
  who   = c(80, 120, 200, 300, 400,
            80, 150, 200, 300, 400),
  proposed = c(120, 160, 200, 300, 400,
               125, 150, 200, 300, 400)
) |>
  dplyr::mutate(pct_increase = 100 * (proposed / who - 1))

knitr::kable(
  table4 |>
    dplyr::select(`Weight band (kg)` = band, Formulation = formulation,
                  `WHO dose (mg)` = who, `Proposed dose (mg)` = proposed,
                  `Increase (%)` = pct_increase),
  digits = 2,
  caption = "Palmer 2025 Table 4: WHO and proposed daily moxifloxacin doses."
)

low_band <- table4$pct_increase[table4$band == "5-<7"]
std_7_10 <- table4$pct_increase[table4$band == "7-<10" &
                                  table4$formulation == "Standard 400 mg"]
big_bands <- table4$pct_increase[table4$wt_lo >= 10]

stopifnot(
  # "adjusted upwards by 50-56% in 5-7 kg for both formulations"
  all(low_band >= 50 & low_band <= 56.5),
  # "and 33% in 7-10 kg for standard formulation"
  abs(std_7_10 - 33) < 1,
  # "doses in children >10 kg remained the same as WHO recommended"
  all(big_bands == 0)
)
```

## Steady-state AUC equals dose divided by clearance

The paper derives every simulated exposure from
`AUC0-24,ss = (Dose * Bioavailability) / Clearance` (Methods 2.3). Because
bioavailability is fixed at 1 and the model is linear, a steady-state interval
of the solved ODE system must reproduce that closed form. Both sides use the
same parameter values here (random effects are zeroed), so the only difference
is trapezoidal integration error and the bound can be tight.

```{r closed-form}
typical_events <- function(wt, dose, occ = 2, selfadmin = 0, by = 0.05) {
  dosing <- tibble::tibble(
    id = 1L, time = seq(0, tau * (n_dose - 1), by = tau),
    amt = dose, evid = 1L, cmt = "depot"
  )
  obs <- tibble::tibble(
    id = 1L, time = seq(t_end - tau, t_end, by = by),
    amt = NA_real_, evid = 0L, cmt = "central"
  )
  dplyr::bind_rows(dosing, obs) |>
    dplyr::mutate(WT = wt, OCC = occ, SELFADMIN = selfadmin) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

trapz <- function(x, y) sum(diff(x) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)

wt_check <- 15.6   # cohort median weight, Table 1
dose_check <- 200  # WHO dose for the 10-<16 kg band, Table 4

sim_cf <- rxode2::rxSolve(mod_typical, typical_events(wt_check, dose_check),
                          omega = NA, sigma = NA) |>
  as.data.frame() |>
  dplyr::filter(time >= t_end - tau)

auc_solved <- trapz(sim_cf$time, sim_cf$Cc)
cl_check <- exp(th[["lcl"]]) * (wt_check / 16)^th[["e_wt_cl"]]
auc_closed <- dose_check * exp(th[["lfdepot"]]) / cl_check

cat(sprintf("solved AUC(0-24,ss) = %.4f mg*h/L; Dose/CL = %.4f mg*h/L; ratio = %.5f\n",
            auc_solved, auc_closed, auc_solved / auc_closed))

# Realised ratio 0.99887. Two deterministic sources of the shortfall: the deep
# peripheral compartment (Q/F = 0.860 L/h into Vp/F = 44.5 L) has not fully
# equilibrated after 12 daily doses, and the linear trapezoid under-reads a
# peaked profile. Neither depends on the random draw or the thread count. A
# mis-transcribed CL, reference weight or allometric exponent moves this by tens
# of percent, so 0.5% is still a live gate.
stopifnot(abs(auc_solved / auc_closed - 1) < 0.005)
```

## The transit chain is actually driving absorption

`cl` / `vc` parameter pairs can make rxode2 substitute a solved linear
compartment model and silently discard an explicitly written ODE system. If
that happened here the four-step transit chain would be gone and absorption
would no longer depend on MAT at all. Quadrupling MAT must therefore move Tmax
substantially, and the transit states must appear in the solved output.

```{r transit-live}
mod_slow <- mod |> rxode2::ini(lmat = log(4 * 1.01)) |> rxode2::zeroRe()

tmax_of <- function(m) {
  s <- rxode2::rxSolve(m, typical_events(wt_check, dose_check),
                       omega = NA, sigma = NA) |>
    as.data.frame() |>
    dplyr::filter(time >= t_end - tau)
  list(tmax = s$time[which.max(s$Cc)] - (t_end - tau),
       cmax = max(s$Cc),
       states = names(s))
}

base_p <- tmax_of(mod_typical)
slow_p <- tmax_of(mod_slow)

cat(sprintf("MAT = 1.01 h: Tmax = %.2f h, Cmax = %.3f mg/L\n", base_p$tmax, base_p$cmax))
cat(sprintf("MAT = 4.04 h: Tmax = %.2f h, Cmax = %.3f mg/L\n", slow_p$tmax, slow_p$cmax))

stopifnot(
  # The transit chain exists as ODE states in the solved output.
  all(c("depot", "transit1", "transit2", "transit3", "central", "peripheral1") %in%
        base_p$states),
  # Absorption really is MAT-driven: realised 2.00 h -> 6.20 h.
  slow_p$tmax - base_p$tmax > 2,
  # And slower absorption flattens the peak: realised 3.08 -> 2.32 mg/L.
  slow_p$cmax < base_p$cmax
)
```

Tmax at the published MAT is `r sprintf("%.1f", base_p$tmax)` h, against the
median Tmax of 2.1 h that Palmer 2025 Table 3 reports for the trial
participants.

# Virtual cohort

Original observed data are not publicly available. Two virtual cohorts are used
below, both capped at 200 participants per arm.

1. A **CATALYST-like cohort** for the Table 3 comparison: 200 children with
   weights drawn from a log-normal truncated to the observed 6.9-42.1 kg range
   and centred on the observed median of 15.6 kg, half dosed on the standard and
   half on the dispersible formulation at the WHO weight-band doses of Table 4.
   The paper does not publish the weight distribution beyond its median and
   range, so the log-normal shape is an assumption (see Errata).
2. A **dosing-simulation cohort** replicating the paper's Figure 4: for each WHO
   weight band, weight is drawn uniformly across the band, exactly as Results 3.2
   describes ("Uniformly distributed body weight ... from 5 to 31 kg was applied
   in the population simulation").

Both cohorts use `OCC` values corresponding to directly observed clinic doses,
so `SELFADMIN = 0` and the interoccasion variability is not inflated. The
unobserved-dosing arm is exercised separately in its own section below.

```{r cohort}
# set.seed() seeds R's RNG (used for the weight draws). It does NOT seed
# rxode2's simulation RNG, whose streams are partitioned per solver thread, so
# the etas differ between a 2-thread CI runner and a 16-thread workstation.
# Every assertion below is therefore written to hold for any cohort this model
# can produce.
set.seed(20250523)
n_per_arm <- 200

who_dose <- function(wt, dispersible) {
  dplyr::case_when(
    wt < 7             ~ 80,
    wt < 10            ~ dplyr::if_else(dispersible, 150, 120),
    wt < 16            ~ 200,
    wt < 24            ~ 300,
    TRUE               ~ 400
  )
}

catalyst <- tibble::tibble(
  id = seq_len(n_per_arm),
  WT = pmin(pmax(exp(stats::rnorm(n_per_arm, log(15.6), 0.45)), 6.9), 42.1),
  dispersible = rep(c(FALSE, TRUE), length.out = n_per_arm)
) |>
  dplyr::mutate(
    dose = who_dose(WT, dispersible),
    # OCC 2 = observed standard-formulation dose at the PK1 visit,
    # OCC 4 = observed dispersible-formulation dose at the PK2 visit.
    OCC = dplyr::if_else(dispersible, 4, 2),
    SELFADMIN = 0,
    arm = dplyr::if_else(dispersible, "Scored dispersible tablet", "Nondispersible tablet")
  )

expand_events <- function(subjects, obs_by = 0.1) {
  dosing <- subjects |>
    dplyr::select(id, WT, OCC, SELFADMIN, arm, dose) |>
    tidyr::expand_grid(time = seq(0, tau * (n_dose - 1), by = tau)) |>
    dplyr::mutate(amt = dose, evid = 1L, cmt = "depot") |>
    dplyr::select(-dose)
  obs <- subjects |>
    dplyr::select(id, WT, OCC, SELFADMIN, arm) |>
    tidyr::expand_grid(time = seq(t_end - tau, t_end, by = obs_by)) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
  dplyr::bind_rows(dosing, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

catalyst_events <- expand_events(catalyst)
stopifnot(!anyDuplicated(catalyst_events[, c("id", "time", "evid")]))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(mod, events = catalyst_events, keep = c("WT", "arm")) |>
  as.data.frame()

stopifnot(dplyr::n_distinct(sim$id) == n_per_arm)

# Time after dose within the final, steady-state, dosing interval.
sim_ss <- sim |>
  dplyr::filter(time >= t_end - tau) |>
  dplyr::mutate(tad = time - (t_end - tau))
```

# Replicate published figures

```{r figure-2, fig.cap = "Steady-state concentration-time profiles by formulation, in the style of Figure 2 of Palmer 2025."}
band_profile <- sim_ss |>
  dplyr::group_by(arm, tad) |>
  dplyr::summarise(
    lo = stats::quantile(Cc, 0.05),
    md = stats::quantile(Cc, 0.50),
    hi = stats::quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot(band_profile, aes(tad, md)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~arm) +
  labs(x = "Time after dose (h)", y = "Moxifloxacin concentration (mg/L)",
       title = "Steady-state profiles, WHO weight-band dosing",
       caption = "Median with 5th-95th percentile band. Figure 2 of Palmer 2025 shows the observed VPC separated by formulation.")
```

The two formulations overlay almost exactly, which is the model's structural
consequence of the paper finding no significant formulation effect on either
bioavailability or MAT.

```{r figure-4-cohort}
# Replicates the simulation design of Palmer 2025 Figure 4: for each WHO weight
# band, weight is uniform across the band and the band's dose is taken from
# Table 4, once for the WHO dosing table and once for the proposed one.
set.seed(20250524)
n_band <- 100  # per band x dosing-table x formulation arm

band_cohort <- table4 |>
  dplyr::select(band, formulation, wt_lo, wt_hi, who, proposed) |>
  tidyr::pivot_longer(c(who, proposed), names_to = "dosing_table", values_to = "dose") |>
  dplyr::mutate(dosing_table = dplyr::recode(dosing_table,
                                             who = "WHO", proposed = "Proposed")) |>
  dplyr::mutate(arm_id = dplyr::row_number()) |>
  tidyr::expand_grid(k = seq_len(n_band)) |>
  dplyr::mutate(
    id = (arm_id - 1L) * n_band + k,
    WT = stats::runif(dplyr::n(), wt_lo, wt_hi),
    OCC = 2, SELFADMIN = 0,
    arm = paste(formulation, dosing_table, band, sep = " | ")
  )

band_sim <- rxode2::rxSolve(
  mod, events = expand_events(band_cohort, obs_by = 0.25),
  keep = c("WT", "arm")
) |>
  as.data.frame() |>
  dplyr::filter(time >= t_end - tau)

band_metrics <- band_sim |>
  dplyr::group_by(id, arm) |>
  dplyr::summarise(auc = trapz(time, Cc), cmax = max(Cc), .groups = "drop") |>
  tidyr::separate(arm, into = c("formulation", "dosing_table", "band"),
                  sep = " \\| ") |>
  dplyr::mutate(band = factor(band, levels = levels(table4$band)),
                dosing_table = factor(dosing_table, levels = c("WHO", "Proposed")))
```

```{r figure-4, fig.cap = "Replicates Figure 4 of Palmer 2025.", fig.height = 6, fig.width = 7}
# Aggregated adult reference exposures, Palmer 2025 Figure 4 caption and
# Supplementary Material 6: AUC0-24,ss median 37 (15-76) mg*h/L, Cmax median
# 3.06 (1.36-6.72) mg/L.
ref <- list(auc = c(lo = 15, md = 37, hi = 76),
            cmax = c(lo = 1.36, md = 3.06, hi = 6.72))

band_long <- band_metrics |>
  tidyr::pivot_longer(c(auc, cmax), names_to = "metric", values_to = "value") |>
  dplyr::mutate(metric = dplyr::recode(
    metric,
    auc = "AUC(0-24,ss) (mg*h/L)", cmax = "Cmax (mg/L)"
  ))

ref_df <- tibble::tibble(
  metric = c("AUC(0-24,ss) (mg*h/L)", "Cmax (mg/L)"),
  lo = c(ref$auc[["lo"]], ref$cmax[["lo"]]),
  md = c(ref$auc[["md"]], ref$cmax[["md"]]),
  hi = c(ref$auc[["hi"]], ref$cmax[["hi"]])
)

ggplot(band_long, aes(band, value, fill = dosing_table)) +
  geom_rect(data = ref_df, inherit.aes = FALSE,
            aes(xmin = -Inf, xmax = Inf, ymin = lo, ymax = hi),
            fill = "red", alpha = 0.10) +
  geom_hline(data = ref_df, aes(yintercept = md), colour = "red") +
  geom_boxplot(outlier.size = 0.4, coef = 0) +
  facet_grid(metric ~ formulation, scales = "free_y") +
  labs(x = "WHO weight band (kg)", y = NULL, fill = "Dosing table",
       title = "Simulated steady-state exposures by weight band",
       caption = paste("Red line: aggregated adult reference median.",
                       "Shaded: adult 2.5th-97.5th percentile range.")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

## Gating the paper's dosing conclusions

Palmer 2025 concludes that "WHO-recommended doses yielded AUC0-24,ss similar to
those in adults for children >10 kg, but that the median AUC0-24,ss in the
smallest children were considerably below the adult reference median", and that
the proposed doses were chosen "to produce median AUC0-24,ss closest to the
reference median". Because bioavailability is fixed and the model is linear,
the typical exposure in each band is exactly `Dose / CL(WT)`, so these claims
can be checked deterministically at the band midpoints without relying on a
drawn cohort.

```{r dosing-claims}
band_typical <- table4 |>
  dplyr::mutate(
    wt_mid = (wt_lo + wt_hi) / 2,
    cl = exp(th[["lcl"]]) * (wt_mid / 16)^th[["e_wt_cl"]],
    auc_who = who / cl,
    auc_proposed = proposed / cl,
    pct_who = 100 * (auc_who / ref$auc[["md"]] - 1),
    pct_proposed = 100 * (auc_proposed / ref$auc[["md"]] - 1)
  )

knitr::kable(
  band_typical |>
    dplyr::select(`Weight band (kg)` = band, Formulation = formulation,
                  `Midpoint WT (kg)` = wt_mid,
                  `AUC, WHO dose (mg*h/L)` = auc_who,
                  `WHO vs adult median (%)` = pct_who,
                  `AUC, proposed dose (mg*h/L)` = auc_proposed,
                  `Proposed vs adult median (%)` = pct_proposed),
  digits = 1,
  caption = paste("Typical steady-state AUC0-24 by weight band against the",
                  "aggregated adult reference median of 37 mg*h/L.")
)

stopifnot(
  # Every proposed dose lands the typical child near the adult reference median.
  # Realised range -8.5% to +4.4%; 15% still breaks on a mis-transcribed CL,
  # reference weight or allometric exponent, which move AUC by tens of percent.
  all(abs(band_typical$pct_proposed) < 15),
  # WHO dosing is adequate above 10 kg ...
  all(abs(band_typical$pct_who[band_typical$wt_lo >= 10]) < 15),
  # ... and underdoses below 10 kg. Realised -34.6% (5-<7 kg, both
  # formulations), -24.5% (7-<10 kg standard), -5.6% (7-<10 kg dispersible),
  # which is why the paper proposes a dose increase for the standard
  # formulation in 7-<10 kg but not for the dispersible one.
  all(band_typical$pct_who[band_typical$band == "5-<7"] < -20),
  band_typical$pct_who[band_typical$band == "7-<10" &
                         band_typical$formulation == "Standard 400 mg"] < -15
)
```

# PKNCA validation

```{r pknca}
nca_conc <- sim_ss |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, arm, tad, Cc)

nca_dose <- catalyst |>
  dplyr::mutate(tad = 0) |>
  dplyr::select(id, arm, tad, dose)

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ tad | arm + id,
                             concu = "mg/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(nca_dose, dose ~ tad | arm + id, doseu = "mg")

intervals <- data.frame(
  start = 0, end = tau,
  auclast = TRUE, cmax = TRUE, tmax = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
nca_df <- as.data.frame(nca_res)

stopifnot(nrow(nca_df) > 0, dplyr::n_distinct(nca_df$id) == n_per_arm)
```

# Comparison against published NCA

Palmer 2025 Table 3 reports model-derived steady-state exposure metrics for the
trial participants, grouped by formulation. Those values are individual
empirical-Bayes-based metrics for the 36 enrolled children at their actual
recorded doses, so the comparison below is against a virtual cohort dosed at the
nominal WHO weight-band doses rather than a re-simulation of the trial itself.

```{r nca-comparison}
simulated_nca <- nca_df |>
  dplyr::select(arm, id, PPTESTCD, PPORRES)

published <- tibble::tibble(
  arm = rep(c("Nondispersible tablet", "Scored dispersible tablet"), each = 3),
  PPTESTCD = rep(c("auclast", "cmax", "tmax"), 2),
  # Palmer 2025 Table 3, median column.
  PPORRES = c(35.6, 3.58, 2.1,
              39.9, 4.09, 2.1)
)

comparison <- nlmixr2lib::ncaComparisonTable(
  simulated = simulated_nca,
  reference = published,
  by = "arm",
  params = c("auclast", "cmax", "tmax"),
  units = c(auclast = "mg*h/L", cmax = "mg/L", tmax = "h"),
  tolerance_pct = 20
)

knitr::kable(comparison,
             caption = "Simulated versus Palmer 2025 Table 3 median exposures.")
```

```{r nca-comparison-gate}
# ncaComparisonTable()'s "% diff" column is formatted character, so the gate is
# computed numerically here rather than parsed back out of the table.
sim_median <- simulated_nca |>
  dplyr::group_by(arm, PPTESTCD) |>
  dplyr::summarise(sim = stats::median(PPORRES), .groups = "drop")

gate <- dplyr::inner_join(sim_median, published, by = c("arm", "PPTESTCD")) |>
  dplyr::mutate(pct = 100 * (sim / PPORRES - 1))

knitr::kable(gate, digits = 2,
             caption = "Numeric percentage difference used for the gate.")

stopifnot(
  # Realised across two independently drawn cohorts: AUC -4.2% / -8.7% and
  # -9.2% pooled; Cmax +0.8% / -6.9% and -6.2% pooled; Tmax -4.8% throughout.
  # 25% leaves headroom for the weight-distribution assumption and for the
  # difference between nominal WHO weight-band doses and the trial's actual
  # dosing records, while a mis-transcribed clearance, volume or dose moves
  # these by tens of percent. Do not tighten it back on one run.
  all(abs(gate$pct) < 25)
)

# The paper's central claim is that the two formulations are interchangeable.
# The packaged model has no formulation term, so the arms can differ only
# through their weight-band doses; the published Table 3 medians differ by 12%.
auc_ratio <- gate$sim[gate$PPTESTCD == "auclast" &
                        gate$arm == "Scored dispersible tablet"] /
  gate$sim[gate$PPTESTCD == "auclast" & gate$arm == "Nondispersible tablet"]
cat(sprintf("simulated dispersible / standard AUC ratio = %.3f (published 39.9/35.6 = %.3f)\n",
            auc_ratio, 39.9 / 35.6))
stopifnot(abs(auc_ratio - 1) < 0.20)
```

# Interoccasion variability and unobserved dosing

The one model feature that no exposure summary in the paper exercises is the
1.70-fold inflation of the interoccasion variability on dosing occasions that
were not directly observed. It can nonetheless be read straight back out of a
simulation, because steady-state exposure is exactly `Dose * F / CL`: taking
logs, `sd(log(AUC))` equals `sqrt(omega_CL^2 + (scale * omega_F)^2)` with the
weight held fixed. That expression contains all three quantities at once, so it
is a joint check on `IIV on CL` = 13.6% CV, `IOV on F` = 30.8% CV, and the
1.70 fold change, and it also discriminates the variance-versus-standard-
deviation reading of Table 2 (reading 0.308 as a variance would inflate the
prediction by about 70%).

```{r iov-readback}
set.seed(20250525)

iov_arm <- function(occ, selfadmin, id_offset, label) {
  tibble::tibble(
    id = id_offset + seq_len(n_per_arm),
    WT = 16,            # fixed, so weight contributes no spread
    dose = 200,
    OCC = occ, SELFADMIN = selfadmin, arm = label
  )
}

iov_subjects <- dplyr::bind_rows(
  # OCC 2: the directly observed dose given at the PK1 study visit.
  iov_arm(2, 0, 0L, "Observed dosing occasion"),
  # OCC 1: the predose sample arising from the previous day's at-home dose.
  iov_arm(1, 1, 1000L, "Unobserved dosing occasion")
)

iov_sim <- rxode2::rxSolve(mod, events = expand_events(iov_subjects, obs_by = 0.25),
                           keep = c("WT", "arm")) |>
  as.data.frame() |>
  dplyr::filter(time >= t_end - tau)

iov_auc <- iov_sim |>
  dplyr::group_by(arm, id) |>
  dplyr::summarise(auc = trapz(time, Cc), .groups = "drop") |>
  dplyr::group_by(arm) |>
  dplyr::summarise(sd_log_auc = stats::sd(log(auc)), .groups = "drop")

omega_cl <- 0.136          # Table 2, IIV on CL = 13.6 %CV; footnote c
omega_f  <- 0.308          # Table 2, IOV on F  = 30.8 %CV; footnote c
scale_unobs <- th[["e_selfadmin_iovscale"]]

expected <- tibble::tibble(
  arm = c("Observed dosing occasion", "Unobserved dosing occasion"),
  predicted = c(sqrt(omega_cl^2 + omega_f^2),
                sqrt(omega_cl^2 + (scale_unobs * omega_f)^2))
)

readback <- dplyr::inner_join(iov_auc, expected, by = "arm") |>
  dplyr::mutate(pct = 100 * (sd_log_auc / predicted - 1))

knitr::kable(
  readback |>
    dplyr::rename("Occasion" = arm, "sd(log AUC), simulated" = sd_log_auc,
                  "sd(log AUC), predicted" = predicted,
                  "Difference (%)" = pct),
  digits = 4,
  caption = "Interoccasion-variability read-back from the simulated exposures."
)

ratio_sim <- readback$sd_log_auc[readback$arm == "Unobserved dosing occasion"] /
  readback$sd_log_auc[readback$arm == "Observed dosing occasion"]
ratio_pred <- readback$predicted[readback$arm == "Unobserved dosing occasion"] /
  readback$predicted[readback$arm == "Observed dosing occasion"]

cat(sprintf("sd(log AUC) ratio unobserved/observed: simulated %.3f, predicted %.3f\n",
            ratio_sim, ratio_pred))

stopifnot(
  # Realised across two independently drawn cohorts: -3.0% / +1.4% and
  # -1.2% / -5.1%, on 200-subject arms where the sampling standard error of an
  # SD is about 5%. 20% admits that noise but not a variance-versus-SD
  # misreading of Table 2, which would inflate the prediction by about 70%.
  all(abs(readback$pct) < 20),
  # And the ratio isolates the 1.70 fold change; realised +4.6% and -4.0%.
  abs(ratio_sim / ratio_pred - 1) < 0.20
)
```

# Assumptions and deviations

* **Occasion structure.** Palmer 2025 states that predose samples arising from
  prior doses were treated as a separate occasion from postdose samples, and the
  supplementary NONMEM control stream applies one shared interoccasion variance
  per parameter through an `IF (OCC.EQ.N)` block, but neither states the occasion
  count explicitly. Four occasions are encoded here because the CATALYST design
  gives exactly four per child: PK1 predose (from the previous day's at-home
  dose), the observed PK1 dose, PK2 predose, and the observed PK2 dose.
  Occasions 2-4 are `fixed()` to the occasion-1 variance, which is how nlmixr2
  expresses NONMEM's `$OMEGA BLOCK(1) SAME`.
* **`SELFADMIN` covariate.** The source column is `UNOBS`. It maps onto the
  registered `SELFADMIN` canonical with identical polarity (1 = unobserved /
  self-administered), so no value inversion is needed. Its role here differs
  from the founding Wallender 2021 usage: it scales the interoccasion-variability
  standard deviation on bioavailability and MAT rather than shifting a typical
  value. In the CATALYST design `SELFADMIN = 1` exactly on occasions 1 and 3.
* **Formulation.** The dispersible-versus-standard effects on bioavailability
  (ratio 1.05, 90% CI 0.95-1.15) and on MAT (ratio 1.06, 90% CI 0.90-1.26) were
  estimated but not retained, so the packaged model has no formulation term, in
  line with Table 2. The two arms above therefore differ only through their
  weight-band doses.
* **Screened but unused covariates.** Height-for-age z-score, weight-for-age
  z-score, age (as both a sigmoidal maturation function and a proportional
  effect), fat-free mass as an alternative body-size descriptor, and HIV status
  were all tested and rejected. They are recorded in the model file's
  `covariatesDataExcluded` list with the paper's reasons and are not required to
  simulate this model.
* **Observation count.** The paper is internally inconsistent: Results paragraph
  1 states that 438 observations were obtained, of which 4 predose samples below
  the quantification limit were excluded, while the Table 1 header gives
  `obs = 384` for the analysis dataset. The model file records 384 per Table 1
  and notes the discrepancy.
* **Weight distribution.** Table 1 publishes only the median (15.6 kg) and range
  (6.9-42.1 kg) of body weight, not its shape. The CATALYST-like cohort above
  uses a log-normal on that median, truncated to that range; a different shape
  would shift the pooled Table 3 medians somewhat. The Figure 4 replication does
  not depend on this assumption because the paper specifies uniform weights
  within each band.
* **Table 3 is not a re-simulation of the trial.** Table 3 reports individual
  model-derived metrics for the 36 enrolled children at their actual recorded
  doses, and the paper notes that in some weight bands the milligram dose given
  at PK2 differed from that given at PK1 because the two formulations have
  different tablet sizes. The comparison above uses nominal WHO weight-band
  doses, which is the likeliest source of the residual differences in the
  comparison table. The wider simulated ranges are expected: extreme order
  statistics grow with cohort size, and 200 subjects per arm are simulated
  against 36 enrolled children.
* **Extrapolation below 7 kg.** The paper's own discussion warns that
  model-informed doses in children under 7 kg are extrapolated beyond the
  observed CATALYST weight range and rest on a model with no maturation
  component. The simulations here stay within the 5-31 kg range of Figure 4.
* **No maturation or renal-function term.** A published renal-maturation
  function added as a fixed component worsened the objective function by 13
  points, so none is encoded. The paper's worst-case scenario (Figure S2,
  assuming 100% renal clearance with a maturation function) is not reproduced
  here because the maturation function's parameters live in the cited upstream
  reference rather than in this paper.
