---
title: "SERT inhibition and tremor incidence MBMA (Panday 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SERT inhibition and tremor incidence MBMA (Panday 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7.5, fig.height = 5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Panday_2025_sert_tremor_cavg_mbma"))$reference`
* Description (Cavg fit): `r rxode2::rxode(readModelDb("Panday_2025_sert_tremor_cavg_mbma"))$description`
* Article: <https://doi.org/10.1002/cpt.3696>

The serotonin reuptake transporter (SERT) clears serotonin from the synapse.
Excess extrasynaptic serotonin arising from SERT inhibition produces the
adverse-event spectrum called *serotonin syndrome*, of which tremor is the most
frequently reported manifestation. Panday 2025 asked whether the incidence of
tremor across marketed SERT inhibitors could be predicted from a single
mechanistic quantity: how much of the brain SERT population a drug covers at its
clinical dose.

The analysis has three parts (Panday 2025 Figure 1a):

1. **Literature survey** - collate, per drug and per dose, the reported tremor
   incidence, the steady-state plasma exposure, the plasma and brain unbound
   fractions and brain partitioning, and the in vitro SERT potency.
2. **SERT coverage calculation** - convert the reported total plasma
   concentration into an unbound *brain* concentration, then divide by the
   free-corrected in vitro potency to obtain a unitless *target coverage*.
3. **Emax meta-regression** - relate the observed per-arm tremor incidence to
   that coverage with a three-parameter Emax function, weighted by study-group
   size.

Parts 2 and 3 are what this extraction encodes. Part 1 is data curation, and its
outputs are the covariate columns the model consumes.

# The two model files

The paper reports twelve fits of a single structural model - the cross of two
exposure metrics (Cavg, Cmax), two potency metrics (SERT IC50, SERT K~D~) and
three data selections (all drugs; excluding drugs with active metabolites;
subset to SSRIs and SNRIs). The *structure* is identical across all twelve; only
the `ini()` values and the meaning of the supplied exposure covariate change.

Two model files are shipped, split along the one axis that changes the **input
contract** - which exposure metric the user must supply:

```{r models}
mod_cavg <- readModelDb("Panday_2025_sert_tremor_cavg_mbma")
mod_cmax <- readModelDb("Panday_2025_sert_tremor_cmax_mbma")

rxode2::rxode(mod_cavg)
```

Each file ships the paper's most-confident data selection for its metric
("excluding drugs with active metabolites") with an IC50-based potency. The
other ten parameter sets are tabulated below and are applied with `ini()`; the
potency-metric axis needs no separate file because substituting a K~D~ for an
IC50 changes only the value supplied in `IC50_SERT` and the accompanying
parameter set.

# Population

```{r population}
pop <- rxode2::rxode(readModelDb("Panday_2025_sert_tremor_cavg_mbma"))$population
tibble::tibble(
  Field = names(pop),
  Value = vapply(pop, function(x) paste(as.character(x), collapse = "; "), character(1))
) |>
  knitr::kable()
```

The clinical layer is human: 33 treatment arms totalling 29,677 participants
across 20 SERT inhibitors from five drug classes (Panday 2025 Table S1 and
Figure 1b). The distribution layer is not - brain partitioning parameters come
predominantly from rat and mouse studies, and the potencies are in vitro. That
cross-species integration is what makes the analysis *translational*, and it is
the main source of uncertainty in the predicted coverage.

# Scope of this extraction

Included:

* Equation S1 - unbound steady-state plasma concentration, with the ng/mL to nM
  conversion.
* Equation S2b - unbound steady-state brain concentration via K~p,uu~.
* Equation S8 (main-text Eq. 1) - SERT target coverage.
* Equation S9 (main-text Eq. 2) - the Emax relationship for tremor incidence.

Not included in `model()`, and why:

* **Equation S2a** (`C_br,ss,ub = C_pl,ss * f_u,br * K_p * 1000/MW`). The paper
  states a preference for the S2b route because it "considered `f_u,pl`
  measurements to be more reliable than `f_u,br` measurements". S2a is
  algebraically the same quantity: supply `KPUU_BRAIN = K_p * f_u,br / f_u,pl`
  and the two routes agree exactly. This identity is checked numerically below.
* **Equations S3-S6** (Friden pH-partition correction of brain-homogenate
  `f_u,br`). The equations are printed in the supplement but the compartment
  volumes (`V_ISF`, `V_cyto`, `V_lyso`) and compartmental pH values they require
  are not - they live in the cited Friden et al. reference, which is not among
  the on-disk sources. The *outputs* of the correction are tabulated (Table S2,
  parameter rows `brain fu corrected`), so the correction is an input-curation
  step whose results the model consumes rather than a step the model performs.
  No constant has been substituted from outside the on-disk sources.
* **Equation S7** (free-potency correction for serum-containing assays). See
  Errata - the tabulated corrected potencies could not be reproduced from the
  equation as printed, so the model consumes the already-corrected values.
* **The active-metabolite coverage summation.** For amitriptyline/nortriptyline,
  fluoxetine/norfluoxetine and venlafaxine/O-desmethylvenlafaxine the paper
  computed coverage separately per moiety and added the two. Evaluate the model
  once per moiety and add the resulting `sertcov` values to reproduce that; the
  default parameter set is the fit that *excludes* those drugs.
* **Residual error.** Main-text Eq. 2 carries an additive noise term `eta`, but
  no variance for it is reported anywhere on disk. Per the standing policy
  against inventing variances, the model is typical-value only. There is also no
  between-study random effect in this MBMA - the study-group sizes enter as
  regression *weights*, not as a variance component.

# Source trace

Every equation and every `ini()` value, with its location in the source.

```{r source-trace}
tibble::tribble(
  ~Element,                    ~Source,
  "cplssub = CAV * FU * 1000 / MW",
    "Supplemental Methods Eq. S1",
  "cbrssub = cplssub * KPUU_BRAIN",
    "Supplemental Methods Eq. S2b",
  "sertcov = cbrssub / IC50_SERT",
    "Supplemental Methods Eq. S8; main text Eq. 1",
  "tremor = emax * sertcov / (ec50 + sertcov) + b0",
    "Supplemental Methods Eq. S9; main text Eq. 2",
  "emax = exp(lemax), ec50 = exp(lec50), b0 = exp(le0)",
    "Supplemental Methods, final paragraph: 'parameter = exp(theta_parameter)'",
  "lemax = 1.94 (Cavg model)",
    "Table S2, Cavg,ub,ss / SERT IC50, excluding drugs with active metabolites, weighted estimate (SE 0.229)",
  "lec50 = 0.195 (Cavg model)",
    "Table S2, same row block (SE 0.803)",
  "lemax = 1.84 (Cmax model)",
    "Table S2, Cmax,ub,ss / SERT IC50, excluding drugs with active metabolites, weighted estimate (SE 0.257)",
  "lec50 = 0.300 (Cmax model)",
    "Table S2, same row block (SE 1.27)",
  "le0 = fixed(0.293), both models",
    "Table S2 (theta_B0 = 0.293 in all twelve fits); Results and Supplemental Methods state B0 was SET to the group-size-weighted mean placebo tremor incidence of 1.34%, and exp(0.293) = 1.340",
  "CAV / CMAX covariate values",
    "Table S2, 'plasma steady state total Cavg' / '...Cmax' parameter rows",
  "FU covariate values",
    "Table S2, 'plasma fu' parameter rows",
  "MW covariate values",
    "Table S2, 'MW' parameter rows",
  "KPUU_BRAIN covariate values",
    "Table S2, 'unbound brain to plasma ratio' / 'Kpuu' rows, or 'total brain to plasma ratio' x 'brain fu corrected' / 'plasma fu'",
  "IC50_SERT covariate values",
    "Table S2, 'SERT IC50' / 'SERT IC50 derived' parameter rows (geometric mean where several were found)"
) |>
  knitr::kable()
```

# The twelve reported fits

All twelve log-scale parameter sets from Tables S2 and S4. The two rows marked
`*` are the defaults shipped in the two model files.

```{r all-fits}
# Values are carried as STRINGS so that the number of printed decimals -- and
# therefore the rounding interval each published estimate stands for -- survives
# into the tolerance calculation in Gate 1 below.
fits <- tibble::tribble(
  ~exposure, ~potency, ~selection,                 ~lemax_s, ~se_lemax_s, ~lec50_s, ~se_lec50_s, ~le0_s, ~se_le0_s,
  "Cavg", "IC50", "All drugs",                 "2.01", "0.253", "0.0949", "0.926", "0.293", "0.796",
  "Cavg", "IC50", "Excl. active metabolites*", "1.94", "0.229", "0.195",  "0.803", "0.293", "0.620",
  "Cavg", "IC50", "SSRIs and SNRIs",           "1.91", "0.248", "-0.397", "0.974", "0.293", "0.867",
  "Cmax", "IC50", "All drugs",                 "2.00", "0.229", "0.584",  "0.942", "0.293", "1.03",
  "Cmax", "IC50", "Excl. active metabolites*", "1.84", "0.257", "0.300",  "1.27",  "0.293", "1.20",
  "Cmax", "IC50", "SSRIs and SNRIs",           "1.92", "0.311", "0.218",  "1.18",  "0.293", "1.70",
  "Cavg", "KD",   "All drugs",                 "1.85", "0.212", "1.42",   "1.08",  "0.293", "0.944",
  "Cavg", "KD",   "Excl. active metabolites",  "1.77", "0.201", "1.42",   "1.01",  "0.293", "0.764",
  "Cavg", "KD",   "SSRIs and SNRIs",           "1.87", "0.214", "1.42",   "1.06",  "0.293", "0.858",
  "Cmax", "KD",   "All drugs",                 "1.71", "0.691", "1.42",   "2.34",  "0.293", "3.35",
  "Cmax", "KD",   "Excl. active metabolites",  "1.66", "0.739", "1.42",   "2.71",  "0.293", "3.54",
  "Cmax", "KD",   "SSRIs and SNRIs",           "1.74", "1.1",   "1.42",   "3.32",  "0.293", "5.44"
) |>
  dplyr::mutate(
    lemax = as.numeric(lemax_s), se_lemax = as.numeric(se_lemax_s),
    lec50 = as.numeric(lec50_s), se_lec50 = as.numeric(se_lec50_s),
    le0   = as.numeric(le0_s),   se_le0   = as.numeric(se_le0_s)
  )

fits |>
  dplyr::transmute(
    Exposure = exposure, Potency = potency, `Data selection` = selection,
    `theta_Emax (SE)` = paste0(lemax_s, " (", se_lemax_s, ")"),
    `theta_EC50 (SE)` = paste0(lec50_s, " (", se_lec50_s, ")"),
    `theta_B0 (SE)`   = paste0(le0_s,   " (", se_le0_s,   ")")
  ) |>
  knitr::kable(caption = "Panday 2025 Tables S2 (IC50-based) and S4 (K_D-based).")
```

## Gate 1 - reproduce Tables S3 and S5 from the log-scale estimates

Tables S3 and S5 report the *physical* (back-transformed) Emax, EC50 and B0 for
every fit, computed from the log-scale estimates as
`mean = exp(mu + sigma^2/2)`, `sd = sqrt((exp(sigma^2) - 1) * exp(2*mu + sigma^2))`
and `median = exp(mu)`. Reproducing all 36 published values from the 36
log-scale inputs is an enumerating check over the entire reported parameter
space: a transposed column, a variance-versus-SD mix-up or a missing
back-transform would break it immediately. This check is deterministic - no
seed, no cohort, no simulation - so an exact tolerance is appropriate.

```{r gate-lognormal}
ln_mean   <- function(mu, s) exp(mu + s^2 / 2)
ln_sd     <- function(mu, s) sqrt((exp(s^2) - 1) * exp(2 * mu + s^2))
ln_median <- function(mu, s) exp(mu)

# Half-width of the rounding interval a printed value stands for. With a decimal
# point, precision is the number of decimals ("1.20" is 1.20 +/- 0.005, "1.1" is
# 1.1 +/- 0.05). Without one, Tables S3/S5 print three significant figures
# ("36700" is 36700 +/- 50).
printed_half <- function(txt) {
  has_dot <- grepl(".", txt, fixed = TRUE)
  dec <- ifelse(has_dot, nchar(sub("^[^.]*\\.", "", txt)), 0L)
  v <- as.numeric(txt)
  ifelse(has_dot, 0.5 * 10^(-dec),
         ifelse(v == 0, 0, 0.5 * 10^(floor(log10(abs(v))) - 2)))
}

long <- dplyr::bind_rows(
  dplyr::transmute(fits, exposure, potency, selection, parameter = "Emax",
                   mu_s = lemax_s, se_s = se_lemax_s),
  dplyr::transmute(fits, exposure, potency, selection, parameter = "EC50",
                   mu_s = lec50_s, se_s = se_lec50_s),
  dplyr::transmute(fits, exposure, potency, selection, parameter = "B0",
                   mu_s = le0_s,   se_s = se_le0_s)
)

# ln_mean and ln_sd increase in mu and in s (s >= 0); ln_median depends on mu
# alone. The extremes of each statistic over the rounding box of the two inputs
# are therefore attained at the corners of that box.
recomputed <- long |>
  dplyr::mutate(
    mu = as.numeric(mu_s), se = as.numeric(se_s),
    dmu = printed_half(mu_s), dse = printed_half(se_s),
    se_lo = pmax(se - dse, 0), se_hi = se + dse,
    mean = ln_mean(mu, se),
    mean_lo = ln_mean(mu - dmu, se_lo), mean_hi = ln_mean(mu + dmu, se_hi),
    sd = ln_sd(mu, se),
    sd_lo = ln_sd(mu - dmu, se_lo), sd_hi = ln_sd(mu + dmu, se_hi),
    median = ln_median(mu, se),
    median_lo = exp(mu - dmu), median_hi = exp(mu + dmu)
  )

# Published values, Panday 2025 Tables S3 (IC50) and S5 (K_D), carried as
# strings so each one's own printed precision is preserved.
published <- tibble::tribble(
  ~exposure, ~potency, ~selection,                 ~parameter, ~mean_s,  ~sd_s,   ~median_s,
  "Cavg", "IC50", "All drugs",                 "Emax", "7.73",   "1.98",  "7.48",
  "Cavg", "IC50", "All drugs",                 "EC50", "1.69",   "1.97",  "1.10",
  "Cavg", "IC50", "All drugs",                 "B0",   "1.84",   "1.73",  "1.34",
  "Cavg", "IC50", "Excl. active metabolites*", "Emax", "7.15",   "1.66",  "6.96",
  "Cavg", "IC50", "Excl. active metabolites*", "EC50", "1.68",   "1.60",  "1.22",
  "Cavg", "IC50", "Excl. active metabolites*", "B0",   "1.62",   "1.11",  "1.34",
  "Cavg", "IC50", "SSRIs and SNRIs",           "Emax", "6.97",   "1.75",  "6.76",
  "Cavg", "IC50", "SSRIs and SNRIs",           "EC50", "1.08",   "1.36",  "0.67",
  "Cavg", "IC50", "SSRIs and SNRIs",           "B0",   "1.95",   "2.07",  "1.34",
  "Cmax", "IC50", "All drugs",                 "Emax", "7.62",   "1.77",  "7.42",
  "Cmax", "IC50", "All drugs",                 "EC50", "2.79",   "3.34",  "1.79",
  "Cmax", "IC50", "All drugs",                 "B0",   "2.29",   "3.16",  "1.34",
  "Cmax", "IC50", "Excl. active metabolites*", "Emax", "6.52",   "1.70",  "6.31",
  "Cmax", "IC50", "Excl. active metabolites*", "EC50", "3.04",   "6.12",  "1.35",
  "Cmax", "IC50", "Excl. active metabolites*", "B0",   "2.75",   "4.93",  "1.34",
  "Cmax", "IC50", "SSRIs and SNRIs",           "Emax", "7.18",   "2.29",  "6.84",
  "Cmax", "IC50", "SSRIs and SNRIs",           "EC50", "2.50",   "4.34",  "1.24",
  "Cmax", "IC50", "SSRIs and SNRIs",           "B0",   "5.73",   "23.8",  "1.34",
  "Cavg", "KD",   "All drugs",                 "Emax", "6.52",   "1.4",   "6.38",
  "Cavg", "KD",   "All drugs",                 "EC50", "7.41",   "11.0",  "4.15",
  "Cavg", "KD",   "All drugs",                 "B0",   "2.09",   "2.51",  "1.34",
  "Cavg", "KD",   "Excl. active metabolites",  "Emax", "6.01",   "1.22",  "5.89",
  "Cavg", "KD",   "Excl. active metabolites",  "EC50", "6.94",   "9.29",  "4.15",
  "Cavg", "KD",   "Excl. active metabolites",  "B0",   "1.79",   "1.60",  "1.34",
  "Cavg", "KD",   "SSRIs and SNRIs",           "Emax", "6.66",   "1.44",  "6.51",
  "Cavg", "KD",   "SSRIs and SNRIs",           "EC50", "7.27",   "10.45", "4.15",
  "Cavg", "KD",   "SSRIs and SNRIs",           "B0",   "1.94",   "2.02",  "1.34",
  "Cmax", "KD",   "All drugs",                 "Emax", "7.00",   "5.48",  "5.51",
  "Cmax", "KD",   "All drugs",                 "EC50", "64.6",   "1000",  "4.15",
  "Cmax", "KD",   "All drugs",                 "B0",   "372",    "10000", "1.34",
  "Cmax", "KD",   "Excl. active metabolites",  "Emax", "6.91",   "5.88",  "5.26",
  "Cmax", "KD",   "Excl. active metabolites",  "EC50", "165",    "6530",  "4.15",
  "Cmax", "KD",   "Excl. active metabolites",  "B0",   "702",    "36700", "1.34",
  "Cmax", "KD",   "SSRIs and SNRIs",           "Emax", "10.4",   "15.9",  "5.7",
  "Cmax", "KD",   "SSRIs and SNRIs",           "EC50", "1010",   "24800", "4.15",
  # Table S5 prints this SD as "10E+12"; read as 1e13 it agrees with the formula.
  "Cmax", "KD",   "SSRIs and SNRIs",           "B0",   "365000", "1e13",  "1.34"
)

chk <- dplyr::inner_join(recomputed, published,
                         by = c("exposure", "potency", "selection", "parameter"))
stopifnot(nrow(chk) == 36L)

# A value is consistent when the interval implied by rounding of the two
# log-scale inputs overlaps the interval implied by rounding of the published
# value. No fudge factor: a wrong formula, a variance/SD mix-up or a missing
# back-transform breaks this immediately.
consistent <- function(lo, hi, pub_s) {
  pub <- as.numeric(pub_s); h <- printed_half(pub_s)
  pub + h >= lo & pub - h <= hi
}

chk <- chk |>
  dplyr::mutate(
    ok_mean   = consistent(mean_lo,   mean_hi,   mean_s),
    ok_sd     = consistent(sd_lo,     sd_hi,     sd_s),
    ok_median = consistent(median_lo, median_hi, median_s)
  )

# Four published cells are NOT consistent, all in the Cmax / K_D block of
# Table S5 and all low by a factor of about ten (see Errata). Asserting the
# discrepant set EXACTLY -- rather than loosening the tolerance or dropping the
# cells -- keeps the check honest in both directions: a new disagreement fails,
# and so does a disagreement that silently goes away.
discrepant <- chk |>
  dplyr::filter(!ok_mean | !ok_sd | !ok_median) |>
  dplyr::transmute(cell = paste(exposure, potency, selection, parameter,
                                ifelse(!ok_mean, "mean", "sd"))) |>
  dplyr::pull(cell) |>
  sort()

expected_discrepant <- sort(c(
  "Cmax KD All drugs B0 sd",
  "Cmax KD Excl. active metabolites B0 sd",
  "Cmax KD SSRIs and SNRIs EC50 sd",
  "Cmax KD SSRIs and SNRIs B0 mean"
))
stopifnot(identical(discrepant, expected_discrepant))

n_ok <- sum(chk$ok_mean) + sum(chk$ok_sd) + sum(chk$ok_median)
sprintf(paste("Gate 1: %d of %d published values across Tables S3 and S5 are",
              "consistent with the Table S2 / S4 log-scale estimates;",
              "the %d exceptions are enumerated above and in Errata."),
        n_ok, 3L * nrow(chk), 3L * nrow(chk) - n_ok)

chk |>
  dplyr::filter(potency == "IC50") |>
  dplyr::transmute(
    Exposure = exposure, `Data selection` = selection, Parameter = parameter,
    `Mean (calc)` = signif(mean, 3),   `Mean (pub)` = mean_s,
    `SD (calc)`   = signif(sd, 3),     `SD (pub)`   = sd_s,
    `Median (calc)` = signif(median, 3), `Median (pub)` = median_s
  ) |>
  knitr::kable(caption = "Gate 1: Table S3 reproduced from the Table S2 log-scale estimates. All 54 IC50-based values pass; the K_D half (Tables S4/S5) is checked in the same chunk and carries the four documented exceptions.")
```

All 54 values of Table S3 - its 18 rows times mean, standard deviation and
median, covering every IC50-based fit and so every fit either shipped model uses
- reproduce. The four exceptions all sit in Table S5's Cmax / K~D~ block,
which the paper itself identifies as its least well determined analysis, and none
of them touches a parameter used by the extracted models.

## Gate 2 - B0 is the weighted placebo incidence

The paper states in two places that B0 was *set* to the group-size-weighted mean
placebo tremor incidence of 1.34%, not estimated. `exp(theta_B0)` must therefore
return exactly that number, and it must be identical across all twelve fits.

```{r gate-b0}
b0_all <- exp(fits$le0)
stopifnot(
  length(unique(fits$le0)) == 1L,
  abs(round(b0_all[1], 2) - 1.34) < 1e-12,
  abs(b0_all[1] - 1.34) < 5e-4
)
sprintf("exp(theta_B0) = exp(%.3f) = %.4f%% (published weighted placebo mean: 1.34%%)",
        fits$le0[1], b0_all[1])
```

## Gate 3 - the two imputation correlations

Two power-law correlations were fitted to the curated dataset and used to fill
gaps. Each has one published worked answer, which pins the coefficient, the
exponent and the direction of the fit simultaneously.

* `f_u,brain` vs `f_u,plasma` (Figure S2): `y = 0.1495 * x^1.2589`. Applied to
  desvenlafaxine's reported `f_u,plasma` of 0.7 it must return the 0.0954
  recorded in Table S2 (row 848), which the main text rounds to 0.10.
* SERT K~D~ vs SERT IC50 (Figure S4): `y = 0.1625 * x^0.9462`. Applied to the
  free-corrected tapentadol/tramadol SERT IC50 of 3219.5 nM it must return the
  338.78 nM recorded in Table S2 (row 964).

```{r gate-imputation}
impute_fu_brain <- function(fu_plasma) 0.1495 * fu_plasma^1.2589   # Panday 2025 Figure S2
impute_sert_kd  <- function(ic50_nm)   0.1625 * ic50_nm^0.9462     # Panday 2025 Figure S4

fu_br_desven <- impute_fu_brain(0.7)
kd_tapentadol <- impute_sert_kd(3219.5)

stopifnot(
  abs(fu_br_desven  - 0.0954) < 5e-5,   # Table S2 row 848, 3 significant figures
  abs(round(fu_br_desven, 2) - 0.10) < 1e-12,  # main text Results
  abs(kd_tapentadol - 338.78) < 5e-3    # Table S2 row 964, 5 significant figures
)

tibble::tibble(
  Quantity = c("desvenlafaxine f_u,brain from f_u,plasma = 0.7",
               "tapentadol/tramadol SERT K_D from IC50 = 3219.5 nM"),
  Reproduced = c(signif(fu_br_desven, 4), signif(kd_tapentadol, 6)),
  Published  = c(0.0954, 338.78),
  Source     = c("Figure S2; Table S2 row 848", "Figure S4; Table S2 row 964")
) |>
  knitr::kable()
```

# Replicating the fitted relationships (Figure 2 and Figures S3, S5, S6)

Figure 2 of the main text plots tremor incidence against Cavg-based SERT target
coverage for three data selections; Figures S3, S5 and S6 do the same for the
Cmax and K~D~ variants. Below the fitted curves are evaluated through the model
itself by sweeping the covariate inputs so that `sertcov` takes the intended
values.

Because the model computes coverage from its inputs rather than accepting it
directly, the sweep supplies a unit "probe" drug: `FU = 1`, `MW = 1000`,
`KPUU_BRAIN = 1`, `IC50_SERT = 1`, so that `sertcov` reduces to the supplied
concentration. That the algebra collapses this way is itself worth asserting.

```{r probe-identity}
cov_grid <- c(0, 10^seq(-3, 2, length.out = 200))

# Both CAV and CMAX are supplied so the same probe drives either model file;
# each model reads only the exposure column it declares.
probe_events <- function(coverage) {
  data.frame(
    id         = 1L,
    time       = seq_along(coverage) - 1,
    CAV        = coverage,      # ng/mL
    CMAX       = coverage,      # ng/mL
    FU         = 1,
    MW         = 1000,          # so that 1000/MW == 1
    KPUU_BRAIN = 1,
    IC50_SERT  = 1,             # nM
    evid       = 0L
  )
}

sim_probe <- rxode2::rxSolve(
  rxode2::rxode(mod_cavg), events = probe_events(cov_grid),
  keep = c("CAV")
) |>
  as.data.frame()

# The probe really does make sertcov equal the supplied concentration.
stopifnot(max(abs(sim_probe$sertcov - cov_grid)) < 1e-10)
```

```{r figure-2, fig.cap = "Panday 2025 Figure 2 (Cavg / IC50 panels a-c) and Figure S3 (Cmax / IC50): fitted Emax relationships between SERT target coverage and tremor incidence for each data selection. Curves are evaluated through the extracted model; the paper does not tabulate per-arm target coverage, so the observed study points of the published figures cannot be overlaid (see Assumptions)."}
# Dot-prefixed argument names so the ini() non-standard evaluation cannot
# mistake the supplied value for the model parameter of the same name.
# le0 is not overridden: it is fixed(0.293) in every one of the twelve fits.
curve_for <- function(mod, .lemax, .lec50, coverage = cov_grid) {
  m <- rxode2::rxode(mod)
  m <- suppressMessages(rxode2::ini(m, lemax = .lemax, lec50 = .lec50))
  rxode2::rxSolve(m, events = probe_events(coverage)) |>
    as.data.frame() |>
    dplyr::transmute(sertcov, tremor)
}

fits_ic50 <- dplyr::filter(fits, potency == "IC50")

curves <- split(fits_ic50, seq_len(nrow(fits_ic50))) |>
  lapply(function(row) {
    mod <- if (row$exposure == "Cavg") mod_cavg else mod_cmax
    curve_for(mod, row$lemax, row$lec50) |>
      dplyr::mutate(exposure = row$exposure,
                    selection = sub("\\*$", "", row$selection))
  }) |>
  dplyr::bind_rows()

# The coverage grid includes 0 (used by Gate 4 as the y(0) = B0 anchor); drop it
# here only because a log x-axis cannot show it.
ggplot(dplyr::filter(curves, sertcov > 0),
       aes(x = sertcov, y = tremor, colour = selection)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ exposure, ncol = 2,
             labeller = labeller(exposure = c(Cavg = "Cavg,ub,ss / SERT IC50 (Figure 2)",
                                              Cmax = "Cmax,ub,ss / SERT IC50 (Figure S3)"))) +
  scale_x_log10(name = "SERT target coverage, C_br,ss,ub / IC50_SERT (unitless)") +
  scale_y_continuous(name = "Tremor incidence (%)", limits = c(0, NA)) +
  labs(colour = "Data selection",
       caption = "Emax curves from Panday 2025 Table S2; baseline B0 = 1.34% fixed in every fit.") +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Gate 4 - structural anchors of the Emax function

Three points on the curve are known analytically and must hold for every fit.

```{r gate-anchors}
anchor_check <- split(fits_ic50, seq_len(nrow(fits_ic50))) |>
  lapply(function(row) {
    mod <- if (row$exposure == "Cavg") mod_cavg else mod_cmax
    emax <- exp(row$lemax); ec50 <- exp(row$lec50); b0 <- exp(row$le0)
    got <- curve_for(mod, row$lemax, row$lec50,
                     coverage = c(0, ec50, 1e9))
    tibble::tibble(
      exposure = row$exposure, selection = row$selection,
      zero_calc = got$tremor[1],       zero_exp = b0,
      half_calc = got$tremor[2],       half_exp = emax / 2 + b0,
      asym_calc = got$tremor[3],       asym_exp = emax + b0
    )
  }) |>
  dplyr::bind_rows()

stopifnot(
  max(abs(anchor_check$zero_calc - anchor_check$zero_exp)) < 1e-10,
  max(abs(anchor_check$half_calc - anchor_check$half_exp)) < 1e-10,
  # 1e9 coverage is not literally infinite; allow the residual Emax*ec50/1e9 gap.
  max(abs(anchor_check$asym_calc - anchor_check$asym_exp)) < 1e-6
)

anchor_check |>
  dplyr::transmute(
    Exposure = exposure, `Data selection` = selection,
    `y(0) = B0 (%)`            = signif(zero_calc, 4),
    `y(EC50) = Emax/2+B0 (%)`  = signif(half_calc, 4),
    `y(inf) = Emax+B0 (%)`     = signif(asym_calc, 4)
  ) |>
  knitr::kable(caption = "Gate 4: analytical anchors of Eq. S9 recovered from the solved model.")
```

## Gate 5 - the solved model equals the closed form

Both sides use the same parameter values, so the difference here is pure
floating-point error and a tight bound is the right assertion.

```{r gate-closed-form}
closed_form <- function(coverage, lemax, lec50, le0) {
  exp(lemax) * coverage / (exp(lec50) + coverage) + exp(le0)
}

cf <- closed_form(sim_probe$sertcov, 1.94, 0.195, 0.293)
stopifnot(max(abs(sim_probe$tremor - cf)) < 1e-10)
sprintf("Max |solved - closed form| over %d coverage values: %.3g",
        length(cf), max(abs(sim_probe$tremor - cf)))
```

## Gate 6 - the Eq. S2a and Eq. S2b routes agree

The paper allows either brain-exposure route. Supplying `KPUU_BRAIN` directly
(S2b) and supplying `K_p * f_u,br / f_u,pl` in its place (S2a) must give the
same unbound brain concentration and the same predicted incidence. Vortioxetine
is used because Table S2 disambiguates its inputs explicitly ("Used this value
in meta-analysis" against the rat K~p~ of 20.1, and a pH-partition-corrected
brain fraction unbound of 0.0007047).

```{r gate-routes}
vor <- list(
  cmax_ss     = 18,         # ng/mL, steady-state Cmax at 10 mg QD, Table S2
  fu_plasma   = 0.02,       # human, 98% plasma protein binding, Table S2
  mw          = 298.4,      # g/mol, Table S2
  kp          = 20.1,       # total brain:plasma, rat, "Used this value in meta-analysis"
  fu_brain    = 0.0007047,  # pH-partition corrected, Table S2 row 1124
  kpuu_direct = 1.2,        # unbound brain:plasma, wildtype mice, Table S2 row 815
  ic50        = 5.4         # nM, Table S2
)

kpuu_from_kp <- vor$kp * vor$fu_brain / vor$fu_plasma

# The exposure on hand is a steady-state Cmax, so the Cmax model file is the
# right one to drive here and in Gate 7.
route_events <- function(kpuu) {
  data.frame(id = 1L, time = 0, CMAX = vor$cmax_ss, FU = vor$fu_plasma,
             MW = vor$mw, KPUU_BRAIN = kpuu, IC50_SERT = vor$ic50, evid = 0L)
}

s2a <- rxode2::rxSolve(rxode2::rxode(mod_cmax), events = route_events(kpuu_from_kp)) |>
  as.data.frame()

# S2a as the paper writes it: C_br,ss,ub = C_pl,ss * f_u,br * K_p * 1000/MW
s2a_direct <- vor$cmax_ss * vor$fu_brain * vor$kp * 1000 / vor$mw
stopifnot(abs(s2a$cbrssub - s2a_direct) < 1e-12)

sprintf(paste("Eq. S2a written out: %.5g nM; via KPUU_BRAIN = Kp*fu_br/fu_pl = %.5g:",
              "%.5g nM (identical). Kp,uu implied by S2a is %.4f vs %.1f measured directly."),
        s2a_direct, kpuu_from_kp, s2a$cbrssub, kpuu_from_kp, vor$kpuu_direct)
```

The implied K~p,uu~ of `r sprintf("%.3f", 20.1 * 0.0007047 / 0.02)` from the rat
K~p~ route and the mouse-derived direct measurement of 1.2 differ by roughly a
factor of 1.7 for the same drug. This is the cross-species uncertainty the
Discussion acknowledges, and it propagates linearly into the coverage estimate.

## Gate 7 - end-to-end worked example

The full Eq. S1 - S2b - S8 - S9 chain for vortioxetine 10 mg QD, using the K~p~
route Table S2 marks as the one used in the meta-analysis.

```{r worked-example}
sim_vor <- rxode2::rxSolve(rxode2::rxode(mod_cmax),
                           events = data.frame(
                             id = 1L, time = 0, CMAX = vor$cmax_ss, FU = vor$fu_plasma,
                             MW = vor$mw, KPUU_BRAIN = kpuu_from_kp,
                             IC50_SERT = vor$ic50, evid = 0L)) |>
  as.data.frame()

steps <- tibble::tibble(
  Step = c("C_max,ss (input, total plasma)", "Eq. S1: C_pl,ss,ub", "Eq. S2b: C_br,ss,ub",
           "Eq. S8: SERT target coverage", "Eq. S9: predicted tremor incidence"),
  Value = c(vor$cmax_ss, sim_vor$cplssub, sim_vor$cbrssub, sim_vor$sertcov, sim_vor$tremor),
  Units = c("ng/mL", "nM", "nM", "unitless", "%")
)
knitr::kable(steps, digits = 4)
```

The observed vortioxetine 10 mg QD arm reported 0% tremor against a 2.9% placebo
rate in a 100-participant arm (Table S2 rows 832-833); the 5 mg arm reported
4.6% against the same placebo. With a placebo baseline this noisy at n = 100, a
single arm cannot discriminate the fit - which is exactly why the paper weights
by group size and pools 33 arms. The prediction is asserted only to lie inside
the range the model can produce.

```{r gate-worked}
stopifnot(
  sim_vor$tremor > exp(0.293),                          # strictly above baseline B0
  sim_vor$tremor < exp(1.84) + exp(0.293),              # strictly below Emax + B0
  sim_vor$sertcov > 0
)
sprintf("Predicted tremor incidence %.2f%% at a SERT target coverage of %.3f.",
        sim_vor$tremor, sim_vor$sertcov)
```

# Errata and unreproducible source details

* **Equation S7 (free-potency correction) is mis-typeset and could not be
  reproduced.** As printed it reads
  `IC50_SERT = IC50_SERT,m (f_u,pl/(1 - f_u,pl) * r_p + f_u,pl`, with an
  unbalanced parenthesis. Read literally, the multiplier exceeds 1 whenever
  `f_u,pl > 1/(1 + r_p)` - it would make the free potency *weaker* than the
  measured one, the wrong direction for a protein-binding correction. Checked
  against the tabulated corrections in Table S2 it does not reproduce them
  either: l(R)-methadone goes from a measured 0.17 uM in 10% fetal bovine serum
  to a corrected 167.74 nM (a factor of 0.987, whereas the equation with
  `f_u,pl = 0.13`, `r_p = 0.1` gives 0.145), and tapentadol's measured 3.2195 uM
  and corrected 3219.5 nM are the same number. The model therefore consumes the
  already-corrected `IC50_SERT` values from Table S2 as data, and no attempt is
  made to re-derive them.
* **Two Table S2 comment cells transpose their source values.** Row 1099
  (citalopram, `brain fu corrected` = 0.02) says the correction was applied to a
  literature value of 0.0045, and row 1101 (duloxetine, 0.0015606) says 0.06 -
  but row 49 records citalopram's literature brain `f_u` as 0.06 and row 51
  records duloxetine's as 0.0045. The corrected/uncorrected ratios (3.0 and
  2.88) confirm the drug assignment is right and only the two cited literature
  values are swapped between the comments.
* **B0 encoded as `fixed()`.** Table S2 prints a standard error beside
  `theta_B0`, which normally signals estimation. Both the Results section and the
  Supplemental Methods say plainly that the weighted placebo mean of 1.34% "was
  used as" B0, the same 0.293 appears in all twelve fits, and `exp(0.293)` is
  1.340. The bootstrap column, by contrast, varies (-0.0153 to 0.28), consistent
  with the weighted placebo mean being recomputed within each bootstrap
  resample. `fixed()` reflects the point-estimate model as published.
* **No residual-error variance is recoverable.** Main-text Eq. 2 includes an
  additive noise term with no reported variance. The model is typical-value only.
* **The Friden pH-partition constants are not on disk.** See Scope above.
* **Four cells of Table S5 are low by a factor of about ten.** Gate 1
  back-computes all 108 published mean / SD / median values in Tables S3 and S5
  from the log-scale estimates in Tables S2 and S4 using the transformation the
  table captions themselves state, propagating the rounding of each printed
  input. 104 agree. The four that do not are all in the Cmax / K~D~ block of
  Table S5:

  | Fit | Cell | Formula gives | Table S5 prints | Ratio |
  |---|---|---|---|---|
  | Cmax / K~D~ / all drugs | B0 SD | 1.00e5 | 10000 | 10.0 |
  | Cmax / K~D~ / excl. active metabolites | B0 SD | 3.71e5 | 36700 | 10.1 |
  | Cmax / K~D~ / SSRIs and SNRIs | EC50 SD | 2.53e5 | 24800 | 10.2 |
  | Cmax / K~D~ / SSRIs and SNRIs | B0 mean | 3.58e6 | 365000 | 9.8 |

  A consistent factor of ten across four independent cells in one table block,
  with every other cell in the same table reproducing, points to a dropped digit
  in typesetting rather than a different formula. All four sit in the analysis
  the Discussion singles out as least well determined, and none affects either
  shipped model - both are IC50-based, and every Table S3 value reproduces. Gate
  1 asserts this discrepant set *exactly*, so a future correction to the paper
  would make the gate fail loudly rather than pass silently.
* **`10E+12` in Table S5 resolves cleanly.** The B0 standard deviation for the
  Cmax / K~D~ / SSRI-SNRI fit is printed in this unusual form. Read as 1e13 it
  agrees with the formula's 9.54e12 to within the input rounding, so it is a
  notation quirk rather than an error and is checked like any other cell.

# Assumptions and deviations

* **Per-arm target coverage is not tabulated by the paper**, so the observed
  study points of Figures 2, S3, S5 and S6 cannot be placed on the reproduced
  curves. Table S2 supplies the raw ingredients, but several drugs carry multiple
  literature values for the same input (fluoxetine has K~p,uu~ values of 1, 2.6,
  10 and 21.2; vortioxetine has two steady-state Cmax values at 10 mg QD) and
  only some rows are annotated "Used this value". Choosing among them would be
  guesswork, so the vignette validates against the exact published anchors
  instead of reconstructing the scatter. Gate 7 works one arm end to end using
  only rows the sheet disambiguates explicitly.
* **Two files rather than one or twelve.** The structure is identical across all
  twelve reported fits, so a twelve-file split would be noise. A single file was
  rejected because shipping the Cavg parameter set as the default while a user
  supplies Cmax would be silently wrong, and the covariate name is the only place
  that distinction can be enforced.
* **Species mixing.** Tremor incidences and plasma unbound fractions are human;
  brain partitioning is largely rodent; potencies are in vitro. `population$species`
  is recorded as `"human"` because that is the population the *response* was
  observed in, with the cross-species inputs described in `population$notes`.
* **Placebo arms are not summed into `n_subjects`.** The 20,381 placebo records
  in Table S2 double-count placebo groups shared across dose levels of the same
  trial; only the 29,677 treatment-arm participants are totalled.
* **No PKNCA validation.** This model has no PK time course, no dosing event and
  no concentration output - non-compartmental analysis does not apply. The
  algebraic and identity gates above replace it, following the same pattern as
  `Yoshioka_2018_FXa_inhibitors_mbma`.
