---
title: "Palbociclib PK and myelosuppression in children (Panetta 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Palbociclib PK and myelosuppression in children (Panetta 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Panetta et al. (2024) reported the first population pharmacokinetic and
pharmacodynamic analysis of palbociclib in children, using the Pediatric Brain
Tumor Consortium phase 1 study PBTC-042 (NCT02255461). Three models were fitted
and each has its own parameter table, so each is packaged as its own model file:

| Model file | Paper's model | Parameter source | Output |
|---|---|---|---|
| `Panetta_2024_palbociclib` | Population PK, final (AST) model | Table 2, AST column | Palbociclib plasma concentration |
| `Panetta_2024_palbociclib_anc` | Neutrophil PD model | Table 3A (+ Table 2 PK) | ANC |
| `Panetta_2024_palbociclib_plt` | Platelet PD model | Table 3B (+ Table 2 PK) | Platelet count |

The two pharmacodynamic models were estimated separately, with the palbociclib
pharmacokinetics fixed to each individual's post hoc estimates (section 2.4.2),
so they share no parameters with each other. Section 2.4.1 states only that "we
used a similar structural model for platelet dynamics" -- the structure is
common, the numbers are not.

* Article: <https://doi.org/10.3390/pharmaceutics16121528>
* Supplementary material: <https://www.mdpi.com/article/10.3390/pharmaceutics16121528/s1>

```{r models}
mod_pk  <- rxode2::rxode(readModelDb("Panetta_2024_palbociclib"))
mod_anc <- rxode2::rxode(readModelDb("Panetta_2024_palbociclib_anc"))
mod_plt <- rxode2::rxode(readModelDb("Panetta_2024_palbociclib_plt"))
```

* Citation: `r mod_pk$reference`

The PK model runs in hours, matching Table 2. Both PD models run in days,
matching Table 3, and carry the PK layer with its rate constants converted
inline (`log(0.46 * 24)`, `log(36.5 * 24)`, `log(0.8 / 24)`).

## Units and BSA

Panetta 2024 reports `V/F` in L/m^2 and `CL/F` in L/h/m^2, and section 2.3
states that "the palbociclib dosage (i.e., the palbociclib dosage normalized to
the patient body surface area (BSA)) was used as the model input". Doses are
therefore supplied in **mg/m^2**, body-surface area cancels out of the model
entirely, and no BSA covariate is required. `central / vc` is in mg/L, so the
observation equation multiplies by 1000 to give the ng/mL that the paper plots.
A clinical dose in mg becomes `amt = dose / BSA`.

# Population

Thirty-four patients aged 4.9-21.6 years (median 12.8) with recurrent,
progressive or refractory brain tumors were enrolled; one was excluded for
taking phenytoin and omeprazole against protocol, leaving 33 in the PK analysis
and 387 evaluable concentrations. Median weight was 52.5 kg (23.8-110.4) and
median BSA 1.5 m^2 (0.8-2.4); 36% were female and 58% Caucasian. Patients were
split into stratum I (21 patients, not heavily pretreated) and stratum II (12
patients, heavily pretreated). Oral palbociclib was given at 50 (6 patients), 75
(21) or 95 mg/m^2 (6) once daily for 21 days of each 28-day course. Baseline
laboratory medians (range) were albumin 4.2 g/dL (3.2-4.9), AST 25 U/L (11-93),
bilirubin 0.4 mg/dL (0.1-1.2) and creatinine 0.5 mg/dL (0.2-1.2) (Table 1).

Two of the 33 patients had only a single pre-treatment blood count and were
excluded from the pharmacodynamic analysis, which used 190 ANC and 189 platelet
observations from 31 patients in course 1 and 99 of each in course 2.

The same information is available programmatically via each model's
`population` metadata, e.g.
`readModelDb("Panetta_2024_palbociclib")()$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Panetta_2024_palbociclib*.R`. The table
below collects them in one place.

| Equation / parameter | Value | Source location |
|---|---|---|
| `ltlag` (Tlag) | 0.8 h | Table 2, AST column |
| `lka` (ka) | 0.46 /h | Table 2, AST column |
| `lvc` (V/F) | 653 L/m^2 | Table 2, AST column |
| `lcl` (CL/F) | 36.5 L/h/m^2 | Table 2, AST column |
| `e_ast_cl` | -0.3 | Table 2, AST column; p = 0.0066 |
| AST centring value | 25 U/L | Table 1 (cohort median); section 2.3.1 |
| `etaltlag`, `etalka`, `etalvc`, `etalcl` | 0.67, 0.87, 0.04, 0.26 (SD) | Table 2, AST column; footnote "IIV and IOV are reported as a standard deviation" |
| `etaiov_cl_1/2` | 0.12 (SD) | Table 2, AST column |
| `propSd` | 0.32 | Table 2, AST column |
| ANC `lrbase` (Ncirc0) | 3.95 x 10^3/uL | Table 3A |
| ANC `lkin`, `logitfktr`, `lic50` | 0.90 /day, 0.78/0.90, 383.5 | Table 3A |
| ANC `lkout`, `lhill` | 4.80 /day, 1 (both fixed) | Table 3A |
| ANC IIV / IOV | CV% per Table 3A | Table 3A |
| ANC `propSd_ANC` | 0.29 | Table 3A (b) |
| PLT `lrbase` (Ncirc0) | 267.40 x 10^9/L | Table 3B |
| PLT `lkin`, `logitfktr`, `lic50` | 0.93 /day, 0.52/0.93, 559.14 | Table 3B |
| PLT `lkout`, `lhill` | 4.80 /day, 1 (both fixed) | Table 3B |
| PLT `propSd_PLT` | 0.20 | Table 3B (b) |
| One-compartment PK with lag | n/a | Section 2.3 |
| Marrow / transit / circulating ODEs | n/a | Equations 1-5 and Figure S1 |
| `km_fb` (K_M) | derived | Equation 6 |
| Compartment initial conditions | derived | Equation 7 |
| `k_bp = fk_bp * kin`, logit-distributed | n/a | Sections 2.4.1 and 2.4.2 |

## Equation 1 as printed

Equation 1 is printed as

```
dN_PBM/dt = [ kin*K_M/(K_M + N_circ) * IC50^n/(IC50^n + C^n) - k_bp ] * N_PBM
```

The bracket multiplies `N_PBM`, which is what makes the proliferation term
first-order. That reading is forced by the paper's own Equations 6 and 7:
setting `dN_PBM/dt = 0` in the drug-free steady state gives
`kin*K_M/(K_M + N_circ0) = k_bp`, which rearranges to exactly
`K_M = k_bp/(kin - k_bp) * N_circ0` (Equation 6), and the resulting
steady-state states are exactly Equation 7. A zero-order reading of the
proliferation term would instead require `kin > k_out * N_circ0`, i.e.
`0.90 > 18.96`, which is false, and would contradict the paper's stated
positivity condition `kin > k_bp`.

The check below confirms the encoded system sits exactly at steady state with
no drug on board -- a necessary condition for Equations 1, 6 and 7 to be
mutually consistent.

Typical-value solves below use the solve-time argument `omega = NA` rather than
`rxode2::zeroRe()`. `zeroRe()` mutates the model object in place, which can
silently strip the between-subject variability from every later simulation that
reuses it; `omega = NA` affects only the one solve.

```{r steady-state}
drug_free <- function(model, output) {
  ev <- rxode2::et(seq(0, 56, by = 1)) |> as.data.frame()
  ev$AST <- 25
  ev$OCC <- 1
  s <- rxode2::rxSolve(model, ev, returnType = "data.frame", omega = NA)
  range(s[[output]])
}
drug_free(mod_anc, "ANC")
drug_free(mod_plt, "PLT")
```

# Virtual cohort

Original observed data are not publicly available. The simulations below use
virtual cohorts of 200 subjects per dose arm at the cohort-median AST of
25 U/L, dosed once daily for 21 days of a 28-day course as in the paper's own
simulation study (section 2.4.3). `OCC` switches from 1 to 2 at the start of
course 2 so that the inter-occasion random effects change between courses.

```{r cohort}
set.seed(20241128)

pd_events <- function(dose) {
  ev <- rxode2::et(amt = dose, cmt = "depot", ii = 1, until = 20) |>
    rxode2::et(seq(0, 56, by = 0.5), cmt = "central") |>
    as.data.frame()
  ev$AST <- 25
  ev$OCC <- ifelse(ev$time < 28, 1, 2)
  ev
}

simulate_pd <- function(model, output, doses = c(50, 75, 95), n = 200) {
  purrr_free <- lapply(doses, function(d) {
    s <- rxode2::rxSolve(model, pd_events(d), nSub = n,
                         returnType = "data.frame", addDosing = FALSE)
    stopifnot(length(unique(s$sim.id)) == n)
    data.frame(dose = d, id = s$sim.id, time = s$time, value = s[[output]])
  })
  dplyr::bind_rows(purrr_free) |>
    dplyr::filter(!is.na(value)) |>
    dplyr::mutate(dose_label = factor(paste0(dose, " mg/m^2"),
                                      levels = paste0(doses, " mg/m^2")))
}

sim_anc <- simulate_pd(mod_anc, "ANC")
sim_plt <- simulate_pd(mod_plt, "PLT")

# Guard against a silently variability-free cohort. If the between-subject
# random effects were ever stripped (see the note above on zeroRe), every
# subject in an arm would return an identical profile and the nadir CV would be
# exactly zero, with no error raised anywhere.
nadir_cv <- function(x) {
  x |>
    dplyr::group_by(dose_label, id) |>
    dplyr::summarise(nadir = min(value), .groups = "drop") |>
    dplyr::group_by(dose_label) |>
    dplyr::summarise(cv = sd(nadir) / mean(nadir), .groups = "drop")
}
stopifnot(all(nadir_cv(sim_anc)$cv > 0.1), all(nadir_cv(sim_plt)$cv > 0.05))
```

# Replicating Figure 4

Figure 4 of Panetta 2024 plots the simulated median (black), 25th-75th
percentiles (blue band) and 5th-95th percentiles (grey band) of ANC and platelet
count over 56 days for each of the three dosages, with the 21-day dosing
interval marked. Panels A-C are ANC and panels D-F are platelets.

```{r figure4-summaries}
pd_bands <- function(x) {
  x |>
    dplyr::group_by(dose_label, time) |>
    dplyr::summarise(
      p05 = quantile(value, 0.05), p25 = quantile(value, 0.25),
      med = median(value),
      p75 = quantile(value, 0.75), p95 = quantile(value, 0.95),
      .groups = "drop"
    )
}
band_anc <- pd_bands(sim_anc)
band_plt <- pd_bands(sim_plt)
```

```{r figure4-anc, fig.cap = "Replicates Figure 4A-C of Panetta 2024: simulated ANC over 56 days after 21 days of daily palbociclib. Horizontal lines are the grade 2, 3 and 4 neutropenia thresholds (1.5, 1.0 and 0.5 x 10^3/uL)."}
ggplot(band_anc, aes(time)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "grey80") +
  geom_ribbon(aes(ymin = p25, ymax = p75), fill = "steelblue", alpha = 0.5) +
  geom_line(aes(y = med), linewidth = 0.8) +
  geom_hline(yintercept = 1.5, linetype = "solid", colour = "black") +
  geom_hline(yintercept = 1.0, linetype = "dashed", colour = "darkgreen") +
  geom_hline(yintercept = 0.5, linetype = "dotted", colour = "red") +
  annotate("rect", xmin = 0, xmax = 21, ymin = -0.35, ymax = -0.05, fill = "blue") +
  facet_wrap(~dose_label) +
  coord_cartesian(ylim = c(-0.4, 8)) +
  labs(x = "Time (days)", y = "ANC (10^3/uL)")
```

```{r figure4-plt, fig.cap = "Replicates Figure 4D-F of Panetta 2024: simulated platelet count over 56 days. Horizontal lines are the grade 1, 2 and 3 thrombocytopenia thresholds (100, 50 and 25 x 10^9/L)."}
ggplot(band_plt, aes(time)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "grey80") +
  geom_ribbon(aes(ymin = p25, ymax = p75), fill = "steelblue", alpha = 0.5) +
  geom_line(aes(y = med), linewidth = 0.8) +
  geom_hline(yintercept = 100, linetype = "solid", colour = "black") +
  geom_hline(yintercept = 50, linetype = "dashed", colour = "darkgreen") +
  geom_hline(yintercept = 25, linetype = "dotted", colour = "red") +
  annotate("rect", xmin = 0, xmax = 21, ymin = -18, ymax = -3, fill = "blue") +
  facet_wrap(~dose_label) +
  coord_cartesian(ylim = c(-20, 500)) +
  labs(x = "Time (days)", y = "PLT (10^9/L)")
```

The simulated median curves reproduce the published shapes. ANC declines through
the dosing period to a nadir a few days after the last dose (typical-value nadir
at day 24.0, 24.5 and 25.0 for 50, 75 and 95 mg/m^2, against a last dose on day
20), then recovers -- fully by around day 50 at 50 mg/m^2, but only to about 94%
and 85% of baseline by day 56 at 75 and 95 mg/m^2, so recovery is still
incomplete at the end of the 28-day course at the higher dosages. The platelet
count reaches its nadir near the end of dosing (day 18.5-19.5) and rebounds
*above* baseline before settling, peaking at 114-127% of baseline -- the
overshoot visible in Figure 4D-F, which the bounded feedback term generates.

# Replicating Table 4

Table 4 tabulates the proportion of simulated individuals whose nadir falls
below each toxicity threshold. Panetta 2024 sampled from the 31 study patients'
individual conditional distributions; the cohorts here are drawn from the
population distribution instead, which widens the tails (see Assumptions).

```{r table4}
nadir_summary <- function(x, thresholds) {
  x |>
    dplyr::group_by(dose_label, id) |>
    dplyr::summarise(nadir = min(value), .groups = "drop_last") |>
    dplyr::group_by(dose_label) |>
    dplyr::summarise(
      `Median nadir` = round(median(nadir), 2),
      !!paste0("% < ", thresholds[1]) := round(100 * mean(nadir < thresholds[1]), 1),
      !!paste0("% < ", thresholds[2]) := round(100 * mean(nadir < thresholds[2]), 1),
      !!paste0("% < ", thresholds[3]) := round(100 * mean(nadir < thresholds[3]), 1),
      .groups = "drop"
    ) |>
    dplyr::rename(Dosage = dose_label)
}

knitr::kable(nadir_summary(sim_anc, c(1.5, 1.0, 0.5)),
             caption = "Simulated ANC nadir (10^3/uL). Panetta 2024 Table 4 reports % below 1.5 / 1.0 / 0.5 of 45 / 19 / 3.1 at 50 mg/m^2, 72 / 49 / 13 at 75 mg/m^2 and 79 / 69 / 32 at 95 mg/m^2.")

knitr::kable(nadir_summary(sim_plt, c(100, 50, 25)),
             caption = "Simulated platelet nadir (10^9/L). Panetta 2024 Table 4 reports % below 100 / 50 / 25 of 6 / 0 / 0 at 50 mg/m^2, 8 / 6 / 0 at 75 mg/m^2 and 12 / 6 / 3 at 95 mg/m^2.")
```

Read these against Table 4 with the sampling difference in mind; the agreement
is good for central tendency and deliberately imperfect in the tails.

* **Platelets agree closely.** The simulated median nadirs (187, 155 and
  138 x 10^9/L) sit just below the published individual-fit median of
  169 x 10^9/L (section 3.5), and the proportion below 100 x 10^9/L is the
  right order of magnitude at every dosage (1, 14 and 21% against 6, 8 and
  12%).
* **ANC agrees on the median but over-predicts the lower tail.** The simulated
  median nadir at 50 mg/m^2 (1.01 x 10^3/uL) matches the published
  individual-fit median of 1.0 (section 3.4), but the simulated proportion
  below 0.5 x 10^3/uL is 25.5% against Table 4's 3.1%, and the deeper dosages
  are shifted the same way.

Two things drive the tail difference, and both are consequences of simulating a
new virtual cohort rather than re-simulating the study patients:

1. Panetta 2024 sampled each of the 31 patients' *conditional* distributions.
   Those are shrunk toward the typical value, so their spread is narrower than a
   draw from the population distribution by construction.
2. These models retain the population PK random effects, and they are very
   large -- `ka` has an IIV standard deviation of 0.87 on the log scale and
   `Tlag` 0.67. The paper instead fixed each patient's PK to their post hoc
   estimates. Because the drug effect on proliferation saturates asymmetrically,
   a wider exposure distribution deepens the lower percentiles more than it
   raises the upper ones.

Neither is a discrepancy in the encoded parameters: the typical-value nadirs in
the next section land where the paper's own median curves do.

# The IC50 unit discrepancy

Table 3 and the Equation 1-5 glossary label `IC50` and `C_palbociclib` as
**nM**, but the tabulated values are only self-consistent with the paper's own
PK model and simulations when read as **ng/mL** -- the unit the PK model
actually produces. Palbociclib is C24H29N7O2, relative molecular mass
447.54 g/mol, so applying the conversion would multiply the driving
concentration by 2.2344.

Three independent checks, all failed by the nM reading:

1. Supplementary Figure S2 plots each patient's model-estimated concentration on
   a log axis against a left axis *labelled* nM, and the curves peak around
   10^2. This model's steady-state Cmax at 75 mg/m^2 is 121 ng/mL but 271 nM, so
   the plotted numbers are the ng/mL values -- consistent with the paper having
   carried its concentrations in ng/mL throughout while labelling them nM.
2. Both pharmacodynamic models reproduce Figure 4 at all three dosages under the
   ng/mL reading and badly over-predict depletion under the nM reading.
3. Table 4 reports only 6, 8 and 12% of individuals dropping below
   100 x 10^9/L, and only 3.1, 13 and 32% of individuals below
   0.5 x 10^3/uL. Both require typical nadirs well above those thresholds; the
   nM reading puts the typical subject below them at every dosage.

The check below settles it. Both endpoints are simulated at their typical values
(no random effects) under each reading of the tabulated `IC50`.

Note the direction of the conversion. The model is driven by `Cc` in ng/mL, so
reading the tabulated number as nM means the drug is **more** potent on that
scale: `559.14 nM` is only `559.14 / 2.2344 = 250.2 ng/mL`. Dividing `IC50` by
2.2344 is exactly equivalent to multiplying the driving concentration by 2.2344,
and it predicts *deeper* depletion than the packaged reading.

```{r ic50-units}
# nM per (ng/mL) for palbociclib, C24H29N7O2, relative molecular mass 447.54.
MW_FACTOR <- 1000 / 447.54

typical_nadir <- function(model, output, ic50_ngml) {
  vapply(c(50, 75, 95), function(d) {
    s <- rxode2::rxSolve(model, pd_events(d), returnType = "data.frame",
                         omega = NA, params = c(lic50 = log(ic50_ngml)))
    min(s[[output]], na.rm = TRUE)
  }, numeric(1))
}

ic50_check <- data.frame(
  Dosage = c("50 mg/m^2", "75 mg/m^2", "95 mg/m^2"),
  `PLT, published curve` = c(205, 175, 160),
  `PLT, ng/mL (packaged)` = round(typical_nadir(mod_plt, "PLT", 559.14), 0),
  `PLT, nM reading` = round(typical_nadir(mod_plt, "PLT", 559.14 / MW_FACTOR), 0),
  `ANC, ng/mL (packaged)` = round(typical_nadir(mod_anc, "ANC", 383.5), 2),
  `ANC, nM reading` = round(typical_nadir(mod_anc, "ANC", 383.5 / MW_FACTOR), 2),
  check.names = FALSE
)
knitr::kable(ic50_check, caption = "Typical-value nadirs under each reading of the tabulated IC50. PLT in 10^9/L, ANC in 10^3/uL. The PLT reference column is read off the median curves of Figure 4D-F.")
```

Both endpoints reject the nM reading, and neither is ambiguous:

* **Platelets.** The packaged reading tracks the published median curves; the nM
  reading is roughly 40% low at every dosage and puts the typical subject below
  the 100 x 10^9/L threshold that Table 4 says only 6, 8 and 12% of individuals
  cross.
* **Neutrophils.** The packaged reading gives typical nadirs that straddle the
  Table 4 thresholds in the right places (above 1.0 x 10^3/uL at 50 mg/m^2,
  where only 19% of individuals fall below it). The nM reading essentially
  wipes out the circulating pool at every dosage, which is irreconcilable with
  Table 4 reporting only 3.1, 13 and 32% of individuals below 0.5 x 10^3/uL.

The models therefore use the tabulated numbers with the units read as ng/mL.

# Pharmacokinetics and NCA validation

The PK model is validated against the study's own sampling design: a single
oral dose on course 1 day 1 sampled pre-dose and at 0.5, 1, 2, 4, 8, 24 and
48 h, and steady state on course 1 day 21 sampled pre-dose and at 1, 2, 4, 8
and 24 h (section 2.2).

```{r pk-sim}
set.seed(20241128L)

pk_single <- rxode2::et(amt = 75, cmt = "depot") |>
  rxode2::et(c(0, 0.5, 1, 2, 4, 8, 24, 48), cmt = "central") |>
  as.data.frame()
pk_single$AST <- 25
pk_single$OCC <- 1

pk_ss <- rxode2::et(amt = 75, cmt = "depot", ii = 24, until = 24 * 20) |>
  rxode2::et(24 * 20 + c(0, 1, 2, 4, 8, 24), cmt = "central") |>
  as.data.frame()
pk_ss$AST <- 25
pk_ss$OCC <- 2

sim_one <- function(ev, treatment, t0) {
  s <- rxode2::rxSolve(mod_pk, ev, nSub = 200, returnType = "data.frame",
                       addDosing = FALSE)
  stopifnot(length(unique(s$sim.id)) == 200L)
  data.frame(id = s$sim.id, time = s$time - t0, Cc = s$Cc, treatment = treatment) |>
    dplyr::filter(!is.na(Cc))
}

pk_conc <- dplyr::bind_rows(
  sim_one(pk_single, "Day 1 (single dose)", 0),
  sim_one(pk_ss, "Day 21 (steady state)", 24 * 20)
)

pk_dose <- pk_conc |>
  dplyr::distinct(id, treatment) |>
  dplyr::mutate(time = 0, amount = 75)

# Same guard as for the pharmacodynamic cohorts: the simulated exposures must
# actually vary between subjects.
stopifnot(
  pk_conc |>
    dplyr::group_by(treatment, id) |>
    dplyr::summarise(cmax = max(Cc), .groups = "drop_last") |>
    dplyr::summarise(cv = sd(cmax) / mean(cmax), .groups = "drop") |>
    dplyr::pull(cv) |> min() > 0.1
)
```

```{r pk-profile, fig.cap = "Simulated palbociclib plasma concentrations at 75 mg/m^2. Compare with the prediction-corrected visual predictive checks in Figure 2e (single dose) and 2f (steady state) of Panetta 2024."}
pk_conc |>
  # The pre-dose record is exactly zero on day 1 and would drop out of a log
  # axis. It is dropped for plotting only; the PKNCA input below keeps it,
  # because the time-zero row is what anchors the AUC interval at 0.
  dplyr::filter(Cc > 0) |>
  dplyr::group_by(treatment, time) |>
  dplyr::summarise(p05 = quantile(Cc, 0.05), med = median(Cc),
                   p95 = quantile(Cc, 0.95), .groups = "drop") |>
  ggplot(aes(time)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "grey80") +
  geom_line(aes(y = med), linewidth = 0.8) +
  facet_wrap(~treatment, scales = "free_x") +
  scale_y_log10() +
  labs(x = "Time after dose (h)", y = "Palbociclib (ng/mL)")
```

```{r pknca}
conc_obj <- PKNCA::PKNCAconc(pk_conc, Cc ~ time | treatment + id,
                             concu = "ng/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(pk_dose, amount ~ time | treatment + id,
                             doseu = "mg/m^2")

# Only Cmax / Tmax / AUClast are requested here. The study's sampling design is
# too sparse to support lambda-z on a per-subject basis -- the day-21 arm has
# six samples, and in subjects with a long lag or slow absorption only one or
# two of them fall after Tmax -- so asking for half.life / AUCinf per subject
# yields warnings rather than estimates. Those parameters are validated on the
# dense typical-value profile below instead.
intervals <- data.frame(
  start = 0,
  end = c(48, 24),
  treatment = c("Day 1 (single dose)", "Day 21 (steady state)"),
  cmax = TRUE, tmax = TRUE, auclast = TRUE
)

nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_summary <- as.data.frame(nca) |>
  dplyr::filter(!is.na(PPORRES)) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(PPORRES = median(PPORRES), .groups = "drop")
```

`cl.obs` needs an explicit unit conversion before it can be compared with the
paper. Doses are in mg/m^2 and concentrations in ng/mL, so PKNCA returns
`dose / AUC` in mg/m^2 per ng*h/mL; multiplying by 1000 gives L/h/m^2.

```{r nca-cl-factor}
# 1 (mg/m^2)/(ng*h/mL) = 1e6 ng/m^2 / (ng*h/mL) = 1e6 mL/(h*m^2) = 1000 L/(h*m^2)
CL_UNIT_FACTOR <- 1000
```

## Comparison against published values

Panetta 2024 does not tabulate NCA parameters of its own, so the reference
column combines the model's own structural parameters with the pediatric values
the paper cites. The paper quotes a Tmax of 5-8 h and a half-life of
11.3-19.5 h from the pediatric phase 1 report (Introduction), and Raetz's
pediatric non-compartmental CL/F of 40.4 L/h/m^2 (section 4).

The clearance check is run on the **typical-value** profile over a dense grid
rather than on the population median. That makes it a closed-form identity:
non-compartmental `dose / AUC(0-inf)` must return the model's own `CL/F` of
36.5 L/h/m^2 exactly. The population median cannot serve this purpose here,
because the very large inter-individual variability on `ka` (SD 0.87 on the log
scale) drives some subjects into flip-flop kinetics, where the terminal slope
reflects absorption rather than elimination and `AUC(0-inf)` is biased low.

```{r pknca-typical}
pk_typ <- rxode2::et(amt = 75, cmt = "depot") |>
  rxode2::et(seq(0, 168, by = 0.25), cmt = "central") |>
  as.data.frame()
pk_typ$AST <- 25
pk_typ$OCC <- 1

typ <- rxode2::rxSolve(mod_pk, pk_typ, returnType = "data.frame", omega = NA)
typ_conc <- data.frame(id = 1L, treatment = "Typical value",
                       time = typ$time, Cc = typ$Cc) |>
  dplyr::filter(!is.na(Cc))
typ_dose <- data.frame(id = 1L, treatment = "Typical value",
                       time = 0, amount = 75)

typ_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(typ_conc, Cc ~ time | treatment + id,
                   concu = "ng/mL", timeu = "h"),
  PKNCA::PKNCAdose(typ_dose, amount ~ time | treatment + id,
                   doseu = "mg/m^2"),
  intervals = data.frame(start = 0, end = 168, treatment = "Typical value",
                         cmax = TRUE, tmax = TRUE, auclast = TRUE,
                         half.life = TRUE, aucinf.obs = TRUE, cl.obs = TRUE)
))

typ_wide <- as.data.frame(typ_nca) |>
  dplyr::filter(!is.na(PPORRES)) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::mutate(cl.obs = cl.obs * CL_UNIT_FACTOR)

# Closed-form identity: AUC(0-inf) must equal Dose / (CL/F).
stopifnot(abs(typ_wide$aucinf.obs - 75 / 36.5 * 1000) < 1)
```

```{r nca-comparison}
reference <- data.frame(
  treatment = "Typical value",
  tmax = 6.5,
  half.life = 15.4,
  cl.obs = 36.5
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = dplyr::select(typ_wide, treatment, tmax, half.life, cl.obs),
  reference = reference,
  by = "treatment",
  params = c("tmax", "half.life", "cl.obs"),
  units = c(tmax = "h", half.life = "h", cl.obs = "L/h/m^2")
)
knitr::kable(cmp, caption = "Typical-value single-dose NCA versus reference values. Tmax and half-life references are the midpoints of the 5-8 h and 11.3-19.5 h ranges quoted in the Introduction of Panetta 2024; CL/F is the model's own Table 2 estimate.")
attr(cmp, "footnote")
```

The population summary below is descriptive rather than a validation target, for
the flip-flop reason given above.

```{r nca-table}
nca_summary |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::mutate(dplyr::across(dplyr::any_of("cl.obs"),
                              \(x) x * CL_UNIT_FACTOR)) |>
  dplyr::select(dplyr::any_of(c("treatment", "cmax", "tmax", "auclast",
                                "aucinf.obs", "half.life", "cl.obs"))) |>
  dplyr::rename(Treatment = treatment) |>
  knitr::kable(digits = 2,
               caption = "Median simulated NCA parameters at 75 mg/m^2 (200 subjects per arm). AUC is in ng*h/mL, concentrations in ng/mL, Tmax and half-life in h, clearance in L/h/m^2.")
```

The AST covariate is checked directly against the two values quoted in section
3.2, which are the only covariate predictions the paper prints.

```{r ast-covariate}
data.frame(
  `AST (U/L)` = c(10, 25, 100),
  `Published CL/F` = c(48.2, NA, 24.1),
  `Model CL/F` = round(36.5 * (c(10, 25, 100) / 25)^-0.3, 2),
  check.names = FALSE
) |>
  knitr::kable(caption = "Apparent oral clearance versus AST (L/h/m^2). Panetta 2024 section 3.2 reports 48.2 at AST = 10 U/L and 24.1 at AST = 100 U/L; 25 U/L is the cohort median at which the power model is centred.")
```

# Assumptions and deviations

* **IC50 units.** Table 3 and the Equation 1-5 glossary label `IC50` and
  `C_palbociclib` as nM. The packaged models read the tabulated values as
  **ng/mL**, because the nM reading contradicts Supplementary Figure S2,
  Figure 4 and Table 4 of the same paper. See "The IC50 unit discrepancy"
  above for the three checks. This is the single largest interpretive decision
  in this extraction and it changes the predicted depth of myelosuppression by
  roughly a factor of two.
* **The missing `N_PBM` factor in Equation 1.** The proliferation term is
  first-order in `N_PBM`. This is forced by Equations 6 and 7, not assumed; see
  "Equation 1 as printed" above.
* **`k_bp` parameterisation.** Table 3 tabulates `k_bp` (0.78 and 0.52 /day),
  but sections 2.4.1 and 2.4.2 state that what was actually estimated is the
  fraction `fk_bp = k_bp / kin`, on a logit scale, so that `kin > k_bp` always
  holds. The models encode `logitfktr = qlogis(k_bp / kin)` and recover
  `k_bp` as `fk_bp * kin`. This matters for simulation: with independent
  log-normal random effects on `kin` and `k_bp`, roughly a third of simulated
  subjects would draw `kin < k_bp` and get a negative feedback constant `K_M`.
* **`fk_bp` variability scale.** Table 3 reports the `k_bp` IIV and IOV as CV%,
  but a logit-normal parameter has no CV-to-variance identity. The logit-scale
  standard deviations were obtained with the delta method,
  `sd_logit = CV / (1 - p)`, and each is round-trip checked in the model file
  against the published CV. Both are small and poorly estimated in the source
  (RSE 99-203%).
* **Occasion definition.** Panetta 2024 reports inter-occasion variability on
  CL/F (PK, two sampling visits) and on `kin`, `k_bp` and `IC50` (PD, courses 1
  and 2) but never names an occasion column. The canonical `OCC` covariate is
  used, with two occasions. In the PD models the PK inter-occasion term is keyed
  to the same course index; the PK analysis's own occasions were the course 1
  day 1 and day 21 visits, about three weeks apart and so comparable in span to
  a treatment course. Set `OCC = 1` throughout for a single-occasion simulation.
* **Base versus final PK model.** The Abstract quotes the base-model estimates
  (V/F 664.5 L/m^2, CL/F 36.8 L/h/m^2, ka 0.48 /h). The packaged models use the
  AST column of Table 2, which is the final model that Figure 2's diagnostics
  and the bootstrap refer to.
* **PK residual error in the PD models.** The PD models were fitted with the
  pharmacokinetics fixed to individual post hoc estimates, so their likelihood
  contains only the blood-count residual error. `Cc` is therefore exposed in the
  PD models as a derived variable with no error model; the plasma residual error
  lives in `Panetta_2024_palbociclib`.
* **PK inter-individual variability retained in the PD models.** The PD fit used
  each patient's own post hoc PK parameters. For simulation the population PK
  random effects are retained so that a virtual cohort has realistic exposure
  variability.
* **Simulation cohorts.** Panetta 2024 simulated by sampling each of the 31
  study patients' conditional distributions (n = 10 per patient). The cohorts
  here are drawn from the population distribution, which has wider tails because
  it is not shrunk toward the typical value, and they retain the large population
  PK random effects that the paper replaced with individual post hoc estimates.
  Central tendency agrees closely (see "Replicating Table 4"), but the simulated
  ANC lower tail is materially heavier than Table 4 -- 25.5% below
  0.5 x 10^3/uL at 50 mg/m^2 against the published 3.1%. Treat the tail
  percentiles of this vignette as a property of the virtual cohort, not as a
  reproduction of Table 4.
* **NCA reference values.** Panetta 2024 reports no NCA table of its own. The
  Tmax and half-life references are midpoints of the 5-8 h and 11.3-19.5 h
  ranges quoted in the Introduction from the pediatric phase 1 report, and are
  therefore approximate.
* **No BSA covariate.** Doses are in mg/m^2 and volumes and clearances are per
  m^2, so BSA cancels exactly. This reproduces the paper's own model input
  convention rather than adding a scaling covariate.
