---
title: "Isoniazid, rifampicin, pyrazinamide and ethambutol in children with tuberculous meningitis (Panjasawatwong 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Isoniazid, rifampicin, pyrazinamide and ethambutol in children with tuberculous meningitis (Panjasawatwong 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  error = FALSE
)
```

```{r libraries, message = FALSE}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(ggplot2)

rxode2::rxSetSeed(20201216)
set.seed(20201216)
```

Panjasawatwong et al. (2020) described the plasma and cerebrospinal fluid (CSF)
pharmacokinetics of the four first-line antituberculosis drugs in 100
Vietnamese children with tuberculous meningitis (TBM). Each drug was modelled
separately, so the package ships four model files that share this vignette:

```{r load_models}
inh <- nlmixr2lib::readModelDb("Panjasawatwong_2020_isoniazid")
rif <- nlmixr2lib::readModelDb("Panjasawatwong_2020_rifampicin")
pza <- nlmixr2lib::readModelDb("Panjasawatwong_2020_pyrazinamide")
emb <- nlmixr2lib::readModelDb("Panjasawatwong_2020_ethambutol")
```

Reference: Panjasawatwong N, Wattanakul T, Hoglund RM, Bang ND, Pouplin T,
Nosoongnoen W, Ngo VN, Day JN, Tarning J (2020). Population pharmacokinetic
properties of antituberculosis drugs in Vietnamese children with tuberculous
meningitis. *Antimicrob Agents Chemother* 65(1):e00487-20.
[doi:10.1128/AAC.00487-20](https://doi.org/10.1128/AAC.00487-20)

## Population

One hundred children younger than 15 years with suspected TBM were enrolled at
Pham Ngoc Thach Hospital, Ho Chi Minh City, between October 2009 and March 2011
and treated with the 2006 WHO paediatric regimen: isoniazid 5 mg/kg, rifampicin
10 mg/kg, pyrazinamide 25 mg/kg and ethambutol 15 mg/kg once daily (plus
streptomycin for 2 months and adjunctive dexamethasone). Six plasma samples were
drawn per child over days 1, 14, 30 and 90, with two CSF samples on days 30 and
90. Ethambutol could not be quantified in CSF.

| Characteristic (Table 1) | Value |
|---|---|
| Boys / girls | 56 / 44 |
| Body weight, kg | 10.9 (4.0-43) |
| Age, years | 3.0 (0.167-15.0) |
| Weight-for-age z-score | -1.93 (-5.52 to 1.99) |
| TBM grade I / II / III | 58 / 24 / 18 |
| HIV positive / negative / unknown | 4 / 92 / 4 |
| NAT2 fast / intermediate / slow / unknown | 17 / 47 / 28 / 8 |
| CSF protein, g/L | 1.20 (0.1-5) |

Values are median (range) or counts.

## Model structure

All four models share one skeleton (Figure 2 of the paper): oral dose into a
depot, a chain of transit compartments in which the absorption rate constant
equals the transit rate constant (`ktr = (n + 1) / MTT` with `n` = 2 for
isoniazid, rifampicin and ethambutol and `n` = 3 for pyrazinamide), and one or
two disposition compartments. Isoniazid, rifampicin and pyrazinamide carry a
CSF compartment whose volume is fixed from age (Methods Eq 1). Drug enters it
at `QCSF * fu * PC * Cc` and leaves at `QCSF * Ccsf`, so at steady state the
CSF-to-plasma AUC ratio equals `fu * PC`.

Every clearance and inter-compartmental flow is scaled by `(WT / 10.9)^0.75`
and every plasma volume by `WT / 10.9`. Clearance also carries a sigmoidal
maturation function of postmenstrual age, `PMA^HILL / (PMA^HILL + MAT50^HILL)`,
with PMA in months equal to postnatal age plus 9.33 months. The tabulated
estimates are "scaled to typical patient at 10.9 kg and 3 years" (Tables 2-5,
footnote b), so the maturation term is divided by its value at 3 years
(PMA 45.33 months).

Drug-specific terms:

* **Isoniazid:** two-compartment disposition; CL/F is 56.4% lower in NAT2 slow
  acetylators. Fast, intermediate and ungenotyped children form a single
  reference group.
* **Rifampicin:** autoinduction through an enzyme-turnover pool with kENZ,
  Emax and EC50 fixed to the adult values of Smythe et al. (2012). The pool
  multiplies the pre-induced CL/F. The penetration multiplier rises
  exponentially with CSF protein: `PC = 0.844 * exp(0.245 * CSF_TPRO)`.
* **Pyrazinamide:** CL/F changes by +4.76% and Vc/F by -4.65% per unit of
  weight-for-age z-score, and relative bioavailability carries interoccasion
  variability across the four sampling days.
* **Ethambutol:** two-compartment disposition, plasma only; the maturation
  Hill coefficient is fixed to 1.

Residual error is additive on log-transformed concentrations for every matrix
(`lnorm()`).

## Source trace

| Quantity | Isoniazid | Rifampicin | Pyrazinamide | Ethambutol | Source |
|---|---|---|---|---|---|
| CL/F (L/h) | 9.43 | 3.22 (pre-induced) | 1.07 | 28.2 | Tables 2-5 |
| Vc/F (L) | 3.78 | 12.3 | 7.38 | 98.6 | Tables 2-5 |
| Q/F (L/h) | 28.0 | - | - | 16.9 | Tables 2, 5 |
| Vp/F (L) | 15.3 | - | - | 153 | Tables 2, 5 |
| MTT (h) | 0.878 | 1.25 | 0.457 | 1.8 | Tables 2-5 |
| Transit compartments | 2 | 2 | 3 | 2 | Results, each drug |
| MAT50 (months PMA) | 12.7 | 6.81 | 12.1 | 3.99 | Tables 2-5 |
| HILL | 4.7 | 1.38 | 2.73 | 1 (fixed) | Tables 2-5 |
| QCSF/F (L/h) | 13.7 | 0.00482 | 0.0964 | - | Tables 2-4 |
| fu (fixed) | 0.9 | 0.2 | 0.9 | - | Tables 2-4 |
| PC | 1.65 | 0.844 | 1.02 | - | Tables 2-4 |
| Covariate effects | slow acetylator -56.4% on CL | CSF protein +24.5%/(g/L), exponential on PC | WAZ +4.76% on CL, -4.65% on Vc | - | Tables 2-4 |
| Autoinduction (fixed) | - | kENZ 0.00369 1/h, Emax 1.04, EC50 70.5 ug/L | - | - | Table 3; Eqs 2-3 |
| IIV (%CV) | CL 36.8, Q 101 | CL 19.4, Vc 23.0, MTT 85.0, PC 22.0 | CL 20.0, Vc 18.3, MTT 64.5 | F 20.1, Vc 51.1, MTT 18.4, Vp 96.3 | Tables 2-5 |
| IOV (%CV) | - | - | F 19.0 | - | Table 4 |
| Residual variance, plasma / CSF | 0.474 / 0.170 | 0.513 / 0.309 | 0.0274 / 0.0114 | 0.197 / - | Tables 2-5 |
| Allometric exponents | 0.75 (flows), 1 (volumes), fixed | same | same | same | Methods |
| Maturation function | Eq 4 | Eq 4 | Eq 4 | Eq 4 | Methods |
| CSF volume | Eq 1 | Eq 1 | Eq 1 | - | Methods |
| Enzyme pool | - | Eqs 2-3 | - | - | Methods |

## Structural verification

These checks compare a typical-value solve (all random effects zeroed) against
its own closed form, so both sides use the same parameters and the only
difference is numerical integration error. They are asserted tightly.

```{r gate_helpers}
# Two endpoints (Cc, Ccsf): observation rows nominate the endpoint through dvid.
# dvid = 1 on every observation row returns both Cc and Ccsf as columns.
steady_state_events <- function(amt, n_doses, grid = 0.05, two_endpoints = TRUE) {
  t_last <- (n_doses - 1) * 24
  dose <- data.frame(
    time = 0, amt = amt, evid = 1, cmt = "depot",
    ii = ifelse(n_doses > 1, 24, 0), addl = n_doses - 1
  )
  obs <- data.frame(time = seq(t_last, t_last + 24, by = grid))
  obs$amt <- NA_real_
  obs$evid <- 0
  obs$ii <- 0
  obs$addl <- 0
  if (two_endpoints) {
    dose$dvid <- NA_real_
    obs$cmt <- NA_character_
    obs$dvid <- 1
    keep <- c("time", "amt", "evid", "cmt", "ii", "addl", "dvid")
  } else {
    obs$cmt <- "central"
    keep <- c("time", "amt", "evid", "cmt", "ii", "addl")
  }
  rbind(dose[, keep], obs[, keep])
}

trap <- function(time, conc) {
  sum(diff(time) * (utils::head(conc, -1) + utils::tail(conc, -1)) / 2)
}

typical_profile <- function(model, amt, covariates, n_doses = 21, two_endpoints = TRUE) {
  ev <- steady_state_events(amt, n_doses, two_endpoints = two_endpoints)
  for (nm in names(covariates)) ev[[nm]] <- covariates[[nm]]
  out <- rxode2::rxSolve(rxode2::zeroRe(model), ev, returnType = "data.frame")
  out <- out[!duplicated(out$time), ]
  out$time <- out$time - (n_doses - 1) * 24
  out
}

typical_child <- list(WT = 10.9, AGE = 3)
```

**Isoniazid.** At steady state the amount cleared over one interval equals the
dose, so `CL * AUCtau` must equal the dose for both acetylator groups. The
CSF-to-plasma AUC ratio must equal `fu * PC = 0.9 * 1.65 = 1.485`. The
slow-to-fast plasma AUC ratio must equal `1 / (1 - 0.564)`.

```{r gate_inh}
inh_dose <- 5 * 10.9
inh_fast <- typical_profile(inh, inh_dose, c(typical_child, NAT2_SLOW = 0))
inh_slow <- typical_profile(inh, inh_dose, c(typical_child, NAT2_SLOW = 1))
auc_inh_fast <- trap(inh_fast$time, inh_fast$Cc)
auc_inh_slow <- trap(inh_slow$time, inh_slow$Cc)

stopifnot(
  abs(9.43 * auc_inh_fast / inh_dose - 1) < 0.002,
  abs(9.43 * (1 - 0.564) * auc_inh_slow / inh_dose - 1) < 0.002,
  abs(trap(inh_fast$time, inh_fast$Ccsf) / auc_inh_fast - 0.9 * 1.65) < 0.005,
  abs(auc_inh_slow / auc_inh_fast - 1 / (1 - 0.564)) < 0.005,
  # The explicit ODE system survives: no closed-form linCmt() takeover.
  identical(
    rxode2::rxode2(inh)$state,
    c("depot", "transit1", "transit2", "central", "peripheral1", "csf")
  )
)
c(auc_fast = auc_inh_fast, auc_slow = auc_inh_slow)
```

**Pyrazinamide and ethambutol.** The same mass balance holds at the reference
covariates. For pyrazinamide the reference is WAZ = 0 and the IOV is switched
off (`OCC = 0`). The pyrazinamide CSF ratio must equal `0.9 * 1.02 = 0.918`.

```{r gate_pza_emb}
pza_dose <- 25 * 10.9
pza_typ <- typical_profile(pza, pza_dose, c(typical_child, WAZ = 0, OCC = 0))
auc_pza <- trap(pza_typ$time, pza_typ$Cc)

emb_dose <- 15 * 10.9
emb_typ <- typical_profile(emb, emb_dose, typical_child, two_endpoints = FALSE)
auc_emb <- trap(emb_typ$time, emb_typ$Cc)

stopifnot(
  abs(1.07 * auc_pza / pza_dose - 1) < 0.002,
  abs(trap(pza_typ$time, pza_typ$Ccsf) / auc_pza - 0.9 * 1.02) < 0.005,
  abs(28.2 * auc_emb / emb_dose - 1) < 0.002
)
c(auc_pza = auc_pza, auc_emb = auc_emb)
```

**Rifampicin.** Autoinduction raises clearance over the first weeks, so the
check runs after 60 daily doses; the paper reports that steady state was
reached within 40 days. With the pool at steady state, the dose must equal
`AUCtau` times the time-averaged induced clearance. The time average is
`CLpre` times the mean enzyme amount over the interval. The CSF-to-plasma AUC
ratio must equal `fu * PC(protein) = 0.2 * 0.844 * exp(0.245 * protein)` at
each protein level.

```{r gate_rif}
rif_dose <- 10 * 10.9
rif_ss <- lapply(c(0.2, 1.2, 5), function(p) {
  typical_profile(rif, rif_dose, c(typical_child, CSF_TPRO = p), n_doses = 61)
})
rif_ratio <- vapply(rif_ss, function(d) trap(d$time, d$Ccsf) / trap(d$time, d$Cc), numeric(1))
rif_expected <- 0.2 * 0.844 * exp(0.245 * c(0.2, 1.2, 5))
rif_cl_avg <- trap(rif_ss[[2]]$time, 3.22 * rif_ss[[2]]$enz_pool) / 24
auc_rif_ss <- trap(rif_ss[[2]]$time, rif_ss[[2]]$Cc)

rif_day1 <- typical_profile(rif, rif_dose, c(typical_child, CSF_TPRO = 1.2), n_doses = 1)
auc_rif_day1 <- trap(rif_day1$time, rif_day1$Cc)

stopifnot(
  all(abs(rif_ratio / rif_expected - 1) < 0.01),
  abs(rif_cl_avg * auc_rif_ss / rif_dose - 1) < 0.005,
  # Induction lowers steady-state exposure below the first-dose exposure.
  auc_rif_ss < auc_rif_day1
)
data.frame(
  protein_g_L = c(0.2, 1.2, 5),
  csf_plasma_auc_ratio = signif(rif_ratio, 4),
  expected = signif(rif_expected, 4)
)
c(auc_day1 = auc_rif_day1, auc_ss = auc_rif_ss, induced_cl_over_pre = rif_cl_avg / 3.22)
```

For the typical child, the time-averaged steady-state clearance is
`r round(100 * (rif_cl_avg / 3.22 - 1))`% above the pre-induced value. The paper
reports that "the fully induced clearance rate was 80.1% higher than the
preinduced clearance rate" without saying how that figure was summarised. The
enzyme pool's ceiling is `1 + Emax = 2.04`, and the pool runs below it because
plasma rifampicin falls towards the EC50 late in each dosing interval. The
typical-value first-dose to steady-state plasma AUC ratio is
`r round(auc_rif_day1 / auc_rif_ss, 2)`. For comparison, the ratio of the
cohort's post hoc medians is 1.75: 37.6 mg*h/L on day 1 (Table S1, fully
recovered children) against 21.5 mg*h/L at steady state (Table 3).

## Rifampicin exposure by dose, age and CSF protein (Figure 4)

Figure 4 of the paper shows simulated steady-state rifampicin plasma and CSF
AUC0-24 over doses of 10-60 mg/kg in five age groups, and at CSF protein of
0.2, 1.0 and 5.0 g/L. The typical-value curves below use the Table S5 median
body weight for each age group (6.9, 8.5, 10.5, 14.5 and 20.2 kg). The paper's
box medians at 60 mg/kg in 10-year-olds read off the figure are about 125
mg*h/L (plasma) and about 23, 28 and 75 mg*h/L (CSF at 0.2, 1.0 and 5.0 g/L).

```{r fig4, fig.height = 5}
age_groups <- data.frame(
  age_label = c("6 months", "1 year", "2 years", "5 years", "10 years"),
  AGE = c(0.5, 1, 2, 5, 10),
  WT = c(6.90, 8.50, 10.5, 14.5, 20.2)
)
fig4_grid <- expand.grid(
  age_idx = seq_len(nrow(age_groups)),
  dose_mgkg = c(10, 20, 35, 50, 60),
  protein = c(0.2, 1.0, 5.0)
)
fig4 <- do.call(rbind, lapply(seq_len(nrow(fig4_grid)), function(i) {
  g <- fig4_grid[i, ]
  a <- age_groups[g$age_idx, ]
  d <- typical_profile(
    rif, g$dose_mgkg * a$WT,
    list(WT = a$WT, AGE = a$AGE, CSF_TPRO = g$protein),
    n_doses = 61
  )
  data.frame(
    age_label = a$age_label, dose_mgkg = g$dose_mgkg, protein = g$protein,
    auc_plasma = trap(d$time, d$Cc), auc_csf = trap(d$time, d$Ccsf)
  )
}))
fig4$age_label <- factor(fig4$age_label, levels = age_groups$age_label)

fig4_long <- rbind(
  fig4 |>
    dplyr::filter(protein == 1.0) |>
    dplyr::transmute(age_label, dose_mgkg, panel = "Plasma", auc = auc_plasma),
  fig4 |>
    dplyr::transmute(age_label, dose_mgkg, panel = paste0("CSF, protein ", protein, " g/L"), auc = auc_csf)
)

ggplot(fig4_long, aes(dose_mgkg, auc, colour = age_label)) +
  geom_line() +
  geom_point() +
  facet_wrap(~panel, scales = "free_y") +
  labs(
    x = "Rifampicin dose (mg/kg/day)", y = "Steady-state AUC0-24 (mg*h/L)",
    colour = "Age", caption = "Replicates Figure 4 (top row) of Panjasawatwong 2020, typical values."
  )

fig4_check <- fig4 |> dplyr::filter(age_label == "10 years", dose_mgkg == 60)
fig4_check
stopifnot(
  # The CSF-protein effect is uncentred: at 5 g/L the CSF/plasma ratio is
  # 0.2 * 0.844 * exp(0.245 * 5) = 0.574, matching Figure 4 (about 75/125).
  # Centring on the cohort median of 1.2 g/L would give about 0.43.
  abs(fig4_check$auc_csf[fig4_check$protein == 5] /
    fig4_check$auc_plasma[fig4_check$protein == 5] - 0.574) < 0.01
)
```

## Virtual cohort

The paper's Tables 2-5 report median (range) steady-state exposures derived
from the individual (post hoc) estimates of the study children at the study
doses. A virtual cohort resembling the study population is simulated below at
the same doses. Age is log-normal around the 3-year median, truncated to the
observed 2 months to 15 years. Weight is the Table S5 median weight for age,
interpolated on the log scale, with 15% variation, truncated to the observed
4-43 kg. Truncation redraws values outside the limits rather than clamping
them. Weight-for-age z-scores and CSF protein are drawn around the Table 1
medians within the observed ranges. 28% of children are slow acetylators.

```{r cohort}
n_sub <- 200

draw_truncated <- function(n, draw, lower, upper) {
  out <- draw(n)
  bad <- out < lower | out > upper
  while (any(bad)) {
    out[bad] <- draw(sum(bad))
    bad <- out < lower | out > upper
  }
  out
}

age <- draw_truncated(n_sub, function(n) exp(rnorm(n, log(3), 0.9)), 0.167, 15)
wt_median <- exp(stats::approx(
  log(age_groups$AGE), log(age_groups$WT),
  xout = log(pmin(pmax(age, 0.5), 10))
)$y)
# Beyond the Table S5 age span, extend the growth curve with the end slopes.
slope_lo <- log(8.50 / 6.90) / log(1 / 0.5)
slope_hi <- log(20.2 / 14.5) / log(10 / 5)
wt_median <- ifelse(age < 0.5, 6.90 * (age / 0.5)^slope_lo, wt_median)
wt_median <- ifelse(age > 10, 20.2 * (age / 10)^slope_hi, wt_median)
wt_factor <- draw_truncated(n_sub, function(n) exp(rnorm(n, 0, 0.15)), 0.5, 2)
wt <- wt_median * wt_factor
wt[wt < 4 | wt > 43] <- NA
while (anyNA(wt)) {
  i <- is.na(wt)
  wt[i] <- wt_median[i] * exp(rnorm(sum(i), 0, 0.15))
  wt[wt < 4 | wt > 43] <- NA
}

cohort <- data.frame(
  id = seq_len(n_sub),
  AGE = age,
  WT = wt,
  NAT2_SLOW = as.integer(runif(n_sub) < 0.28),
  WAZ = draw_truncated(n_sub, function(n) rnorm(n, -1.93, 1.4), -5.52, 1.99),
  CSF_TPRO = draw_truncated(n_sub, function(n) exp(rnorm(n, log(1.2), 0.8)), 0.1, 5)
)

stopifnot(
  nrow(cohort) <= 200,
  all(cohort$WT >= 4 & cohort$WT <= 43),
  abs(median(cohort$AGE) / 3 - 1) < 0.25,
  abs(median(cohort$WT) / 10.9 - 1) < 0.25
)
summary(cohort[, c("AGE", "WT", "WAZ", "CSF_TPRO")])
```

```{r simulate}
# Daily doses in mg/kg; the last dosing interval is observed on a 0.25-h grid.
cohort_events <- function(mg_per_kg, n_doses, two_endpoints = TRUE) {
  do.call(rbind, lapply(seq_len(nrow(cohort)), function(i) {
    ev <- steady_state_events(mg_per_kg * cohort$WT[i], n_doses,
      grid = 0.25,
      two_endpoints = two_endpoints
    )
    ev <- cbind(id = cohort$id[i], ev)
    ev$time <- ev$time
    for (nm in c("AGE", "WT", "NAT2_SLOW", "WAZ", "CSF_TPRO")) ev[[nm]] <- cohort[[nm]][i]
    ev
  }))
}

simulate_ss <- function(model, mg_per_kg, n_doses, two_endpoints = TRUE, extra = list()) {
  ev <- cohort_events(mg_per_kg, n_doses, two_endpoints)
  for (nm in names(extra)) ev[[nm]] <- extra[[nm]]
  out <- rxode2::rxSolve(model, ev, returnType = "data.frame")
  t_last <- (n_doses - 1) * 24
  out <- out[out$time >= t_last, ]
  out$time <- out$time - t_last
  out
}

sim_inh <- simulate_ss(inh, 5, 21) |>
  dplyr::select(-dplyr::any_of("NAT2_SLOW")) |>
  dplyr::left_join(cohort[, c("id", "NAT2_SLOW")], by = "id") |>
  dplyr::mutate(arm = ifelse(NAT2_SLOW == 1, "Slow acetylators", "Fast and intermediate"))
sim_rif <- simulate_ss(rif, 10, 61) |> dplyr::mutate(arm = "All")
# Occasion 3 (day 30) carries the steady-state IOV draw on F.
sim_pza <- simulate_ss(pza, 25, 21, extra = list(OCC = 3)) |> dplyr::mutate(arm = "All")
sim_emb <- simulate_ss(emb, 15, 21, two_endpoints = FALSE) |> dplyr::mutate(arm = "All")

stopifnot(
  # Random effects actually varied across subjects.
  dplyr::n_distinct(round(sim_rif$Cc[sim_rif$time == 2], 6)) > 50,
  dplyr::n_distinct(round(sim_emb$Cc[sim_emb$time == 2], 6)) > 50
)
```

```{r vpc_plot, fig.height = 6}
vpc_data <- dplyr::bind_rows(
  sim_inh |> dplyr::transmute(id, time, drug = "Isoniazid", Plasma = Cc, CSF = Ccsf),
  sim_rif |> dplyr::transmute(id, time, drug = "Rifampicin", Plasma = Cc, CSF = Ccsf),
  sim_pza |> dplyr::transmute(id, time, drug = "Pyrazinamide", Plasma = Cc, CSF = Ccsf),
  sim_emb |> dplyr::transmute(id, time, drug = "Ethambutol", Plasma = Cc, CSF = NA_real_)
) |>
  tidyr::pivot_longer(c(Plasma, CSF), names_to = "matrix", values_to = "conc") |>
  dplyr::filter(!is.na(conc)) |>
  dplyr::group_by(drug, matrix, time) |>
  dplyr::summarise(
    p05 = quantile(conc, 0.05), p50 = median(conc), p95 = quantile(conc, 0.95),
    .groups = "drop"
  )

ggplot(vpc_data, aes(time, p50, colour = matrix, fill = matrix)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.2, colour = NA) +
  geom_line() +
  facet_wrap(~drug, scales = "free_y") +
  scale_y_log10() +
  labs(
    x = "Time after dose at steady state (h)", y = "Concentration (mg/L)",
    colour = NULL, fill = NULL,
    caption = "Median and 90% interval of the simulated cohort at the study doses."
  )
```

## PKNCA

`PKNCA` computes AUC0-24, Cmax and Tmax over the steady-state interval for each
drug and matrix. The concentration frame is filtered only on missingness, so the
time-zero record survives.

```{r pknca}
run_nca <- function(sim, conc_col) {
  dat <- sim |>
    dplyr::mutate(conc = .data[[conc_col]], treatment = arm) |>
    dplyr::filter(!is.na(conc)) |>
    dplyr::select(id, treatment, time, conc)
  doses <- dat |>
    dplyr::distinct(id, treatment) |>
    dplyr::mutate(time = 0)
  o_conc <- PKNCA::PKNCAconc(dat, conc ~ time | id / treatment)
  o_dose <- PKNCA::PKNCAdose(doses, ~ time | id + treatment)
  intervals <- data.frame(start = 0, end = 24, auclast = TRUE, cmax = TRUE, tmax = TRUE)
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals))
  as.data.frame(res$result) |>
    dplyr::filter(PPTESTCD %in% c("auclast", "cmax", "tmax"))
}

nca_all <- dplyr::bind_rows(
  run_nca(sim_inh, "Cc") |> dplyr::mutate(drug = "Isoniazid", matrix = "Plasma"),
  run_nca(sim_inh, "Ccsf") |> dplyr::mutate(drug = "Isoniazid", matrix = "CSF"),
  run_nca(sim_rif, "Cc") |> dplyr::mutate(drug = "Rifampicin", matrix = "Plasma"),
  run_nca(sim_rif, "Ccsf") |> dplyr::mutate(drug = "Rifampicin", matrix = "CSF"),
  run_nca(sim_pza, "Cc") |> dplyr::mutate(drug = "Pyrazinamide", matrix = "Plasma"),
  run_nca(sim_pza, "Ccsf") |> dplyr::mutate(drug = "Pyrazinamide", matrix = "CSF"),
  run_nca(sim_emb, "Cc") |> dplyr::mutate(drug = "Ethambutol", matrix = "Plasma")
)

stopifnot(
  nrow(nca_all) > 0,
  !any(is.na(nca_all$PPORRES))
)
```

## Comparison against the published exposures

The reference values are the median steady-state exposures of Tables 2-5.

```{r nca_compare}
simulated_long <- nca_all |>
  dplyr::mutate(group = paste(drug, matrix, treatment, sep = " | ")) |>
  dplyr::select(group, PPTESTCD, PPORRES)

published <- tibble::tribble(
  ~group,                                       ~auclast, ~cmax, ~tmax,
  "Isoniazid | Plasma | Fast and intermediate", 6.35,     2.12,  0.942,
  "Isoniazid | CSF | Fast and intermediate",    9.42,     3.14,  0.955,
  "Isoniazid | Plasma | Slow acetylators",      12.4,     2.41,  1.14,
  "Isoniazid | CSF | Slow acetylators",         18.4,     3.58,  1.15,
  "Rifampicin | Plasma | All",                  21.5,     4.94,  2.13,
  "Rifampicin | CSF | All",                     4.08,     0.253, 10.0,
  "Pyrazinamide | Plasma | All",                288,      42.5,  1.08,
  "Pyrazinamide | CSF | All",                   266,      31.3,  3.06,
  "Ethambutol | Plasma | All",                  8.18,     1.26,  2.59
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = simulated_long,
  reference = published,
  by = "group",
  params = c("auclast", "cmax", "tmax"),
  tolerance_pct = 20
)
knitr::kable(cmp)

auc_cmp <- simulated_long |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::group_by(group) |>
  dplyr::summarise(sim = median(PPORRES), .groups = "drop") |>
  dplyr::inner_join(published[, c("group", "auclast")], by = "group") |>
  dplyr::mutate(ratio = sim / auclast)

auc_checked <- auc_cmp |>
  dplyr::filter(!grepl("Ethambutol", group), group != "Rifampicin | CSF | All")
stopifnot(
  # Isoniazid (both matrices), rifampicin plasma and pyrazinamide (both
  # matrices) AUC medians agree with the published medians. Ethambutol and
  # rifampicin CSF are discussed below.
  nrow(auc_checked) == 7,
  all(abs(auc_checked$ratio - 1) < 0.2)
)
```

The isoniazid and pyrazinamide AUC rows (plasma and CSF) and the rifampicin
plasma AUC row reproduce the published medians. The CSF-to-plasma ratios follow
from `fu * PC`, which the structural gate checks exactly. Some rows differ, and
they are reported here rather than tuned:

* **Ethambutol** simulates about 30% below the published AUC and Cmax medians.
  At steady state AUC0-24 equals `F * Dose / CL` exactly, and ethambutol CL/F
  carries no IIV. A median of 8.18 mg*h/L therefore implies a median dose
  about 40% above the nominal 15 mg/kg at the tabulated CL/F of 28.2 L/h.
  The paper's own adult comparison (114 L/h scaled to 70 kg) confirms that
  tabulated CL/F. The most likely explanation is that the dispensed
  milligram doses (rounded to tablet strengths) exceeded the nominal mg/kg;
  the actual doses are not reported.
* **Pyrazinamide Cmax** runs below the published median while AUC agrees.
  For a 10.9-kg child the tabulated Vc/F of 7.38 L bounds the peak from a
  272.5-mg dose at about 37 mg/L, below the published median of 42.5 mg/L,
  so this row, too, points to actual doses above nominal.
* **Rifampicin CSF AUC** runs about 25% above the published median. The CSF
  exposure depends on each child's CSF protein, and the virtual cohort draws
  protein around the baseline median of 1.20 g/L. The published steady-state
  ratio of CSF to plasma AUC medians (4.08 / 21.5 = 0.19) corresponds to
  `0.2 * 0.844 * exp(0.245 * protein)` at a protein of about 0.5 g/L. The
  CSF samples were taken on days 30 and 90, and CSF protein at those visits is
  not reported; it may have fallen from baseline with treatment.
* **Rifampicin CSF Tmax** is about 6 h in the simulation against a published
  median of 10.0 h (range 5.72-13.9). It depends on QCSF and the transit
  absorption rather than on the CSF volume (varying the volume 4-fold moves
  it by under 0.5 h). The simulated CSF Cmax and AUC agree with the table.

## Assumptions and deviations

* **CSF volume equation (Methods Eq 1).** As typeset, Eq 1 reads
  `150 * [38.78 + age^1.071 * (102.6 - 38.78)] / (age^1.071 + 1.297^1.071)` mL,
  which gives about 8 L for a 3-year-old. Here it is read as a sigmoid in
  percent of the 150-mL adult CSF volume:
  `150 * (38.78 + age^1.071 * (102.6 - 38.78) / (age^1.071 + 1.297^1.071)) / 100`,
  giving 58 mL at birth, 126 mL at 3 years and about 154 mL in adults. This
  reading also reproduces the published rifampicin CSF Cmax (0.257 vs 0.253
  mg/L for the typical child at the cohort-median protein).
* **CSF protein effect is uncentred.** The paper does not print a centring
  value for the exponential CSF-protein effect on rifampicin PC. The uncentred
  form `0.844 * exp(0.245 * protein)` reproduces Figure 4's CSF exposures at
  0.2, 1.0 and 5.0 g/L (see the Figure 4 check above). Centring on the cohort
  median of 1.20 g/L would lower the 5.0-g/L CSF AUC by about 25%.
* **WAZ reference.** The paper does not print a centring value for the linear
  WAZ effects on pyrazinamide CL/F and Vc/F. Here WAZ enters uncentred, so the
  tabulated CL/F and Vc/F apply at WAZ = 0, the growth-standard median. Had the
  authors centred on the cohort median of -1.93, the typical CL/F and Vc/F at
  WAZ = 0 would be about 9% higher and 9% lower, respectively. The comparison
  above cannot separate the two because actual doses are uncertain.
* **Maturation normalisation.** Tables 2-5 give estimates "scaled to typical
  patient at 10.9 kg and 3 years". The maturation term is therefore divided by
  its value at 3 years (PMA 45.33 months), so the tabulated CL/F applies to a
  3-year-old of 10.9 kg.
* **IIV and residual error scales.** Tables 2-5 give IIV and IOV as %CV. They
  are converted to log-scale variances with `omega^2 = log(1 + CV^2)`.
  Residual error is described as "variance of the residual variability,
  incorporated as an additive error on the logarithmic scale" (Tables 2-5,
  footnote a). It is encoded as `lnorm(sqrt(variance))`.
* **Transit chain.** "Two fixed-transit absorption compartments (ka was set to
  be identical to ktr)" together with Figure 2 (dose compartment, transit
  compartments, central) is encoded as depot plus `n` transit compartments
  with `ktr = (n + 1) / MTT`. The typical isoniazid Tmax of 0.96 h matches the
  published median of 0.942 h.
* **CSF mass balance.** Drug moving into the CSF is removed from the central
  compartment, following the arrows of Figure 2. Because the CSF volume is
  about 0.1 L, this has a negligible effect on plasma concentrations.
* **NAT2 phenotype.** Slow acetylators are the only group with a separate
  clearance. Fast, intermediate and ungenotyped children share the reference
  clearance, as in the paper's reduced two-group covariate model.
* **Pyrazinamide IOV.** One shared IOV variance on F is used across the four
  sampling occasions (days 1, 14, 30 and 90). `OCC` values outside 1-4 switch
  the IOV off.
* **Rifampicin autoinduction.** kENZ, Emax and EC50 are fixed to the adult
  values of Smythe et al. (2012), as in the paper; EC50 is converted from
  70.5 ug/L to 0.0705 mg/L. The enzyme pool starts at 1 with production equal
  to kENZ.
* **Pharmacodynamic model.** The paper also reports a Weibull time-to-death
  model stratified by baseline TBM grade (Table S3). It is not part of these
  four PK model files.
* **Virtual cohort.** Age, weight, WAZ and CSF protein are drawn independently
  around the Table 1 medians, and weight follows the Table S5 growth curve. The
  study's individual covariates are not published.
