---
title: "Papachristos_2020_bevacizumab_qss"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Papachristos_2020_bevacizumab_qss}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Papachristos A, Karatza E, Kalofonos H, Karalis V.
  Pharmacogenetics in Model-Based Optimization of Bevacizumab Therapy
  for Metastatic Colorectal Cancer. *Int J Mol Sci.* 2020;21(11):3753.
  doi:[10.3390/ijms21113753](https://doi.org/10.3390/ijms21113753).
* Description: Two-compartment quasi-steady-state target-mediated drug
  disposition (TMDD QSS) model for IV bevacizumab and free VEGF-A in
  adults with metastatic colorectal cancer (mCRC), with allometric
  weight scaling and ICAM-1 / VEGF-A genotype covariate effects
  (Table 2 of Papachristos 2020).
* Modality: Therapeutic anti-VEGF-A monoclonal antibody (humanized
  IgG1, MW ≈ 149 kDa) bound to a soluble target (free VEGF-A, MW
  ≈ 45 kDa back-calculated from the paper's BM0 dual-unit reporting).

This vignette validates the binding-QSS variant (Table 2). Papachristos
2020 also reports a descriptive PK model (Table 1, packaged as
`Papachristos_2020_bevacizumab_pk`) and a PK/PD immediate-response
Imax model (Table 3, packaged as `Papachristos_2020_bevacizumab_pkpd`).

> **TODO (co-equality framing).** The three Papachristos 2020 models are
> CO-EQUAL final models, not a hierarchical base / extended pair. Each
> pursues a different objective: descriptive PK, mechanistic target
> binding (this vignette), or drug-effect characterization (PK/PD).
> Reviewer please confirm this framing and the choice to package all
> three.

# Population

The Papachristos 2020 cohort comprised **46 adult patients with
histologically confirmed metastatic colorectal cancer** (ECOG 0-2)
treated at a single Greek centre. Baseline demographics (paper section
2.1):

- Sex: 61% male, 39% female.
- Median (IQR) weight 74.5 kg (64.15-81.75); median (IQR) age 63 years
  (53-72).
- Race / ethnicity: Greek; not formally tabulated.
- Dosing: 5 mg/kg IV every 2 weeks (76% of patients) or 7.5 mg/kg IV
  every 3 weeks (24%).
- Sample sizes: 156 bevacizumab + 169 free VEGF-A serum
  concentrations.

Two SNPs entered the final binding QSS model: `ICAM-1 rs1799969`
(effect on CL) and `VEGF-A rs699947` (effects on K_ss and BM0).

The same metadata is available programmatically via
`readModelDb("Papachristos_2020_bevacizumab_qss")$population`.

# Source trace

| Parameter (model name)                | Value      | Source                                                   |
|---------------------------------------|------------|----------------------------------------------------------|
| `lcl` (CL, L/day)                     | log(0.344) | Papachristos 2020 Table 2, CL                            |
| `lv1` (V1pop, L)                      | log(5.83)  | Papachristos 2020 Table 2, V1pop                         |
| `lq` (Qpop, L/day)                    | log(0.136) | Papachristos 2020 Table 2, Qpop                          |
| `lv2` (V2pop, L)                      | log(3.17)  | Papachristos 2020 Table 2, V2pop                         |
| `lkout` (kout, 1/day)                 | log(0.116) | Papachristos 2020 Table 2, Koutpop                       |
| `lbm0` (BM0, nM)                      | log(0.0137)| Papachristos 2020 Table 2, BM0pop = 0.0137 nM (= 616.5 ng/L) |
| `lkss` (Kss, nM)                      | log(135)   | Papachristos 2020 Table 2, KSSpop                        |
| `allo_cl` (allometric exponent)       | 1.01       | Papachristos 2020 Table 2, "log(weight/70) on CL"        |
| `e_icam1_rs1799969_cl`                | -0.33      | Papachristos 2020 Table 2, "ICAM-1 rs1799969 mutant on CL" |
| `e_vegfa_rs699947_kss`                | +1.22      | Papachristos 2020 Table 2, "VEGF-A rs699947 mutant on KSS" |
| `e_vegfa_rs699947_bm0`                | -0.851     | Papachristos 2020 Table 2, "VEGF-A rs699947 mutant on BM0" |
| `etalcl` variance                     | 0.09548    | Papachristos 2020 Table 2, ω_CL = 0.309 (var = SD²)      |
| `etalq` variance                      | 0.04040    | Papachristos 2020 Table 2, ω_Q = 0.201                   |
| `cov(etalcl, etalq)`                  | -0.06205   | Papachristos 2020 Table 2, ρ(Q,CL) = -0.999 × 0.309 × 0.201 |
| `etalv1` variance                     | 0.02856    | Papachristos 2020 Table 2, ω_V1 = 0.169                  |
| `etalv2` variance                     | 0.30803    | Papachristos 2020 Table 2, ω_V2 = 0.555                  |
| `etalbm0` variance                    | 0.05760    | Papachristos 2020 Table 2, ω_BM0 = 0.24                  |
| `CcpropSd`                            | 0.253      | Papachristos 2020 Table 2, σ_BEVA                        |
| `CvpropSd`                            | 0.290      | Papachristos 2020 Table 2, σ_VEGF                        |

Equations (paper section 2.3 + Gibiansky et al. 2008 QSS-TMDD form):

\[
\dot A_1 = -\mathrm{CL}\,C_\mathrm{free} - Q\,C_\mathrm{free}
            + Q\,(A_2/V_2) - k_\mathrm{int}\,R_c\,V_1
\]
\[
\dot A_2 = Q\,C_\mathrm{free} - Q\,(A_2/V_2)
\]
\[
\dot{T}_\mathrm{tot} = k_\mathrm{in}
            - k_\mathrm{out}\,T_\mathrm{free}
            - k_\mathrm{int}\,R_c
\]

with the QSS algebraic relation

\[
R_c = \tfrac{1}{2}\!\left(C_\mathrm{tot} + T_\mathrm{tot} + K_{ss}
   - \sqrt{(C_\mathrm{tot} - T_\mathrm{tot})^2
            + 2\,K_{ss}(C_\mathrm{tot} + T_\mathrm{tot}) + K_{ss}^2}\right)
\]

\(C_\mathrm{free} = C_\mathrm{tot} - R_c\),
\(T_\mathrm{free} = T_\mathrm{tot} - R_c\),
\(C_\mathrm{tot} = A_1/V_1\),
\(k_\mathrm{in} = k_\mathrm{out}\,\mathrm{BM_0}\),
\(k_\mathrm{int} = \mathrm{CL}/V_1\) (paper section 2.3).

The bevacizumab observation is total (free + complex) drug in mg/L;
the free-VEGF-A observation is \(T_\mathrm{free}\) in ng/L. Both
residual error models are proportional.

# Errata note (text vs Table)

No formal corrigendum has been identified at the time of extraction
(April 2026). One within-paper inconsistency in sister model 3 (the
PK/PD model, Table 3) is documented in the
`Papachristos_2020_bevacizumab_pkpd` vignette; it does not affect the
binding QSS model presented here.

# Virtual cohort

```{r cohort}
set.seed(2020)
n_subj <- 100

cohort <- tibble(
  ID                  = seq_len(n_subj),
  WT                  = pmin(pmax(rlnorm(n_subj, log(74.5), 0.18), 45), 130),
  SNP_ICAM1_RS1799969 = rbinom(n_subj, 1, 0.20),
  SNP_VEGFA_RS699947  = rbinom(n_subj, 1, 0.52)
)
```

```{r events}
inf_dur     <- 1.5 / 24            # 90-min infusion (days)
n_doses_q2w <- 6
sim_horizon <- 84                  # 12 weeks total follow-up
dose_times  <- seq(0, by = 14, length.out = n_doses_q2w)
obs_times   <- sort(unique(c(0, seq(0.05, sim_horizon, by = 0.5))))

build_events <- function(pop, mgkg, dose_times, regimen_label) {
  amt_per_subject <- pop$WT * mgkg
  d_dose <- pop |>
    dplyr::mutate(amt = amt_per_subject) |>
    tidyr::crossing(time = dose_times) |>
    dplyr::mutate(evid = 1L, cmt = "central", dur = inf_dur,
                  treatment = regimen_label)
  d_obs_cc <- pop |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(amt = 0, evid = 0L, cmt = "Cc", dur = NA_real_,
                  treatment = regimen_label)
  d_obs_cv <- pop |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(amt = 0, evid = 0L, cmt = "Cv", dur = NA_real_,
                  treatment = regimen_label)
  dplyr::bind_rows(d_dose, d_obs_cc, d_obs_cv) |>
    dplyr::rename(id = ID) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    as.data.frame()
}

events <- build_events(cohort, mgkg = 5, dose_times = dose_times,
                       regimen_label = "5 mg/kg Q2W")
```

# Simulation

```{r simulate}
mod <- rxode2::rxode2(readModelDb("Papachristos_2020_bevacizumab_qss"))
sim <- rxode2::rxSolve(mod, events = events,
                       keep = c("treatment"),
                       returnType = "data.frame")

# Endpoint compartments: Cc -> CMT 4, Cv -> CMT 5 (after 3 ODE states)
sim_cc <- sim |> dplyr::filter(CMT == 4)
sim_cv <- sim |> dplyr::filter(CMT == 5)
```

# Replicate published figures

## Bevacizumab concentration-time profile (replicates Supplementary Figure S5A diagnostic)

```{r figure-S5A, fig.width = 7, fig.height = 4.5}
sim_cc_summary <- sim_cc |>
  dplyr::filter(time > 0) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    median = stats::median(Cc, na.rm = TRUE),
    lo     = stats::quantile(Cc, 0.05, na.rm = TRUE),
    hi     = stats::quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(sim_cc_summary, aes(time, median)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, fill = "steelblue") +
  geom_line(linewidth = 1, colour = "steelblue") +
  scale_y_log10() +
  labs(
    x = "Time since first dose (days)",
    y = "Bevacizumab concentration (mg/L, log scale)",
    title = "Simulated bevacizumab PK from binding QSS model (5 mg/kg Q2W)",
    subtitle = paste0("Median and 90% prediction interval, N = ", n_subj),
    caption = "Replicates the diagnostic content of Papachristos 2020 Supplementary Figure S5A."
  ) +
  theme_bw()
```

## Free VEGF-A profile and rebound (replicates Supplementary Figure S5B diagnostic)

```{r figure-S5B, fig.width = 7, fig.height = 4.5}
sim_cv_summary <- sim_cv |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    median = stats::median(Cv, na.rm = TRUE),
    lo     = stats::quantile(Cv, 0.05, na.rm = TRUE),
    hi     = stats::quantile(Cv, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(sim_cv_summary, aes(time, median)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, fill = "darkorange") +
  geom_line(linewidth = 1, colour = "darkorange") +
  geom_hline(yintercept = 616.5, linetype = "dashed", colour = "grey40") +
  labs(
    x = "Time since first dose (days)",
    y = "Free VEGF-A (ng/L)",
    title = "Simulated free VEGF-A from binding QSS model (5 mg/kg Q2W)",
    subtitle = "Median and 90% prediction interval. Dashed line = wild-type baseline (616.5 ng/L).",
    caption = "Replicates the diagnostic content of Papachristos 2020 Supplementary Figure S5B."
  ) +
  theme_bw()
```

## Effect of VEGF-A rs699947 on Kss and baseline (replicates Figure 2C/2D)

Paper section 2.3 highlights two rs699947 effects: lower baseline free
VEGF-A (BM0, Figure 2C) and higher in-vivo affinity (higher K_ss,
Figure 2D). The deterministic typical-value simulation below isolates
the genotype contrast at a fixed 74.5 kg / wild-type-ICAM1 reference
subject.

```{r figure-2CD, fig.width = 7, fig.height = 4}
mod_typical <- rxode2::rxode2(readModelDb("Papachristos_2020_bevacizumab_qss")) |>
  rxode2::zeroRe()

typ_grid <- tidyr::crossing(
  WT                  = 74.5,
  SNP_ICAM1_RS1799969 = 0,
  SNP_VEGFA_RS699947  = c(0, 1)
) |>
  dplyr::mutate(id = dplyr::row_number())

build_typ_events <- function(typ_grid, mgkg, dose_times) {
  d_dose <- typ_grid |>
    dplyr::mutate(amt = WT * mgkg) |>
    tidyr::crossing(time = dose_times) |>
    dplyr::mutate(evid = 1L, cmt = "central", dur = inf_dur)
  d_obs_cc <- typ_grid |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(amt = 0, evid = 0L, cmt = "Cc", dur = NA_real_)
  d_obs_cv <- typ_grid |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(amt = 0, evid = 0L, cmt = "Cv", dur = NA_real_)
  dplyr::bind_rows(d_dose, d_obs_cc, d_obs_cv) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    as.data.frame()
}

typ_events <- build_typ_events(typ_grid, 5, dose_times)
sim_typ <- rxode2::rxSolve(mod_typical, events = typ_events,
                           keep = c("SNP_VEGFA_RS699947"),
                           returnType = "data.frame") |>
  dplyr::mutate(
    geno = ifelse(SNP_VEGFA_RS699947 == 1, "rs699947 mutant", "rs699947 wild-type")
  )

ggplot(sim_typ |> dplyr::filter(CMT == 5),
       aes(time, Cv, colour = geno)) +
  geom_line(linewidth = 1) +
  labs(
    x = "Time since first dose (days)",
    y = "Free VEGF-A (ng/L)",
    colour = "VEGF-A rs699947",
    title = "Effect of rs699947 on free VEGF-A baseline and rebound",
    subtitle = "Typical 74.5 kg, wild-type ICAM-1, 5 mg/kg Q2W. Replicates the time-course basis of Papachristos 2020 Figure 2C-D."
  ) +
  theme_bw()

baseline_compare <- sim_typ |>
  dplyr::filter(CMT == 5, time == 0) |>
  dplyr::select(geno, Cv)
knitr::kable(baseline_compare,
             caption = "Pre-dose baseline free VEGF-A by rs699947 genotype (typical-value simulation, ng/L).")
```

# Steady-state and mass-balance checks

This is a binding-target / turnover model — PKNCA on VEGF-A is not
appropriate (VEGF-A is endogenous, not dosed). The two checks below
verify model correctness:

1. **Pre-dose baseline of free VEGF-A matches BM0.** With no drug
   administered, T_free should equal BM0 = 0.0137 nM × 45000 g/mol
   = 616.5 ng/L for a wild-type rs699947 subject and BM0 ×
   exp(-0.851) = 264 ng/L for a mutant carrier.

```{r baseline-check}
expected_bm0 <- c(
  `rs699947 wild-type` = 0.0137 * 45000,
  `rs699947 mutant`    = 0.0137 * 45000 * exp(-0.851)
)
predose <- sim_typ |>
  dplyr::filter(CMT == 5, time == 0) |>
  dplyr::select(geno, observed_predose = Cv) |>
  dplyr::mutate(expected_bm0 = expected_bm0[geno])
knitr::kable(predose,
             caption = "Pre-dose free VEGF-A: simulated vs algebraic expectation (ng/L).")
```

2. **Long-term recovery to baseline.** With dosing stopped after the
   last simulated infusion, free VEGF-A should rebound to its
   genotype-specific BM0 over a few half-lives of the target turnover
   (1/k_out ≈ 8.6 days).

```{r recovery-check}
last_obs_predose <- sim_typ |>
  dplyr::filter(CMT == 5, time >= 84) |>
  dplyr::group_by(geno) |>
  dplyr::summarise(final_Cv = dplyr::last(Cv), .groups = "drop") |>
  dplyr::mutate(expected_bm0 = expected_bm0[geno],
                pct_recovered = round(100 * final_Cv / expected_bm0, 1))
knitr::kable(last_obs_predose,
             caption = "End-of-simulation free VEGF-A vs genotype-specific BM0 (ng/L). Last dose at day 70; observed at day 84 (~14 d ≈ 1.6 × 1/k_out post-dose).")
```

# Bevacizumab PKNCA validation

For completeness, the same single-dose NCA recipe used for the
descriptive PK model can be applied to the QSS-model bevacizumab
output. The two output streams are unified using `cmt` markers in the
event table; the dose-mass conservation differs slightly from the
descriptive PK model because total drug here includes the bound
complex.

```{r pknca}
suppressPackageStartupMessages(library(PKNCA))

start_ss <- max(dose_times)
end_ss   <- start_ss + 14

conc_df <- sim_cc |>
  dplyr::filter(time >= start_ss, time <= end_ss, !is.na(Cc)) |>
  dplyr::mutate(time_rel = time - start_ss) |>
  dplyr::select(id, treatment, time_rel, Cc)

dose_df <- events |>
  dplyr::filter(treatment == "5 mg/kg Q2W", evid == 1, time == start_ss) |>
  dplyr::mutate(time_rel = 0) |>
  dplyr::select(id, treatment, time_rel, amt)

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time_rel | treatment + id,
                             concu = "mg/L", timeu = "day")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time_rel | treatment + id,
                             doseu = "mg")

intervals <- data.frame(
  start    = 0,
  end      = 14,
  cmax     = TRUE,
  tmax     = TRUE,
  cmin     = TRUE,
  auclast  = TRUE,
  cav      = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- PKNCA::pk.nca(nca_data)
knitr::kable(summary(nca_res),
             caption = "Steady-state NCA for bevacizumab (5 mg/kg Q2W, last simulated interval, binding QSS model).")
```

# Assumptions and deviations

- **Co-equality framing of the three Papachristos 2020 models.** The
  PK / Binding QSS / PK/PD models in Papachristos 2020 are co-equal
  final models, each pursuing a different objective (descriptive PK,
  mechanistic target binding [this vignette], drug-effect
  characterization). They are not a hierarchical base / extended
  pair. All three are packaged separately in nlmixr2lib.
- **Bevacizumab molecular weight (MW_BEV = 149,000 g/mol).** Required
  to convert mg input doses to molar units (nM) for the binding
  equations; not reported numerically in Papachristos 2020. The 149
  kDa value is the standard reference for the marketed antibody and
  is documented as a model assumption in the model file's `model()`
  block.
- **Free VEGF-A effective molecular weight (MW_VEGF ≈ 45,000 g/mol).**
  Back-calculated from the paper's BM0 dual-unit reporting (0.0137 nM
  ≡ 616.5 ng/L → MW = 45 kDa, consistent with the homodimeric
  VEGF-A165 form). Documented as a model assumption.
- **1:1 stoichiometric drug-target binding.** The paper states
  "assuming a 1:1 molecular interaction" (section 2.3). The packaged
  QSS model uses 1:1 binding, consistent with the paper.
- **Bevacizumab observation = total drug (free + complex).** The
  paper's ELISA assay (section 4.4) is not described as a free-drug
  assay; total-drug ELISA is the typical default. Flagged as an
  interpretation assumption.
- **k_int = CL / V1.** Per paper section 2.3: "the elimination
  clearance of the bevacizumab-VEGF-A complex... was set equal to the
  CL of free bevacizumab." Implemented exactly as stated.
- **No errata identified.** A literature search (April 2026) found no
  published corrigendum / erratum to Papachristos 2020.
