---
title: "Parhiz_2024_mRNALNP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parhiz_2024_mRNALNP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(PKNCA)
```

# Model and source

* Citation: Parhiz H, Shuvaev VV, Li Q, et al. Physiologically based
  modeling of LNP-mediated delivery of mRNA in the vascular system.
  *Molecular Therapy: Nucleic Acids*. 2024;35(2):102175.
  doi:10.1016/j.omtn.2024.102175. Open access (CC BY-NC-ND).
* Article: <https://doi.org/10.1016/j.omtn.2024.102175>
* Supplemental information (Document S1, Figures S1-S3 + Tables S1-S4)
  and the ADAPT 5 source code (Data S1, mmc2.zip) are archived with the
  article and on PMC: <https://pmc.ncbi.nlm.nih.gov/articles/PMC10992703/>.

This vignette validates the Parhiz et al. 2024 whole-body PBPK and
luciferase-expression model in mice for the bare (untargeted) LNP
parameterization. The packaged file in
`inst/modeldb/pharmacokinetics/Parhiz_2024_mRNALNP.R` carries the full
24-state ODE system: blood and a blood-cleared LNP pool; vascular pools
for lung, heart, kidney, spleen, liver, portal-organ remainder, and
carcass remainder; and per-tissue intracellular LNP, mRNA, and
luciferase signal compartments for the five major sampled tissues
(lung, heart, kidney, spleen, liver).

# Population

The model was fit to data from male C57BL/6 mice (6-8 weeks old; The
Jackson Laboratory) injected with 8 ug of luciferase-encoding
modified-mRNA loaded into lipid nanoparticles via retro-orbital IV
bolus. n = 3 mice were destructively sampled per time point across
blood and five organs (lung, heart, kidney, spleen, liver). The
physiological parameters (tissue volumes, blood flows) were taken from
the BioDMET database and scaled to a 25-g mouse (paper Table S3); only
the 12 LNP / luciferase parameters in `ini()` were estimated. The
present file packages the bare-LNP parameter set (no surface antibody
conjugation, no active targeting). The vignette below documents the
alternate parameter sets for control-IgG-coated and PECAM-targeted
LNPs (paper Tables 1-3) that share the same structural model.

```{r read-model}
mod <- readModelDb("Parhiz_2024_mRNALNP")
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to
each `ini()` entry in
`inst/modeldb/pharmacokinetics/Parhiz_2024_mRNALNP.R`. The table below
collects them in one place for review.

| Equation / parameter | Value (bare LNP) | Source |
|----------------------|------------------|--------|
| Venous blood mass balance | derivation | Supplement S1 "Pharmacokinetic Model / Venous Blood"; ADAPT 5 XP(1) |
| Tissue vasculature mass balance | derivation | Supplement S1 "Lungs"/"Liver"/"Other Organs / Vasculature"; ADAPT 5 XP(3), XP(5), XP(7), XP(9), XP(11) |
| Intracellular LNP / mRNA / luciferase ODEs | derivation | Supplement S1 "Other Organs / Internalized" and "Pharmacodynamic Model"; ADAPT 5 XP(4), XP(15), XP(20) for lung and analogues |
| `cl_lu` (lung non-specific uptake CL) | 0.134 mL/h | Table 1, Bare LNPs column |
| `cl_he` (heart non-specific uptake CL) | 0.0999 mL/h | Table 1, Bare LNPs column |
| `cl_ki` (kidney non-specific uptake CL) | 0.428 mL/h | Table 1, Bare LNPs column |
| `cl_sp` (spleen non-specific uptake CL) | 0.656 mL/h | Table 1, Bare LNPs column |
| `cl_li` (liver non-specific uptake CL) | 16.3 mL/h | Table 1, Bare LNPs column |
| `cl_bl` (blood elimination CL) | 1.96 mL/h | Table 1, Bare LNPs column |
| `klnp` (intracellular LNP degradation rate k_deg) | 1.00 1/h | Table 1, Bare LNPs column |
| `slu` (lung luciferase production rate S_lung) | 2.57e5 LU/mg protein per ug RNA / h | Table 2, Bare LNPs homogenate column |
| `she` (heart luciferase production rate S_heart) | 6.70e4 LU/mg protein per ug RNA / h | Table 2, Bare LNPs homogenate column |
| `ski` (kidney luciferase production rate S_kidney) | 1.12e5 LU/mg protein per ug RNA / h | Table 2, Bare LNPs homogenate column |
| `ssp` (spleen luciferase production rate S_spleen) | 3.82e5 LU/mg protein per ug RNA / h | Table 2, Bare LNPs homogenate column |
| `sli` (liver luciferase production rate S_liver) | 8.64e5 LU/mg protein per ug RNA / h | Table 2, Bare LNPs homogenate column |
| `kluc` (luciferase decay, non-liver) | 0.0940 1/h | Table 2, Bare LNPs homogenate column |
| `kliv` (luciferase decay, liver) | 0.247 1/h | Table 2, Bare LNPs homogenate column |
| `krna` (mRNA degradation rate, fixed) | 0.114 1/h | Methods text; Anderson 2011 Nucleic Acids Res ref 36 |
| Mouse blood total volume `v_bl` | 1.533 mL | Table S3 |
| Cardiac output `q_co` | 605.4 mL/h | Table S3 (10.1 mL/min x 60) |
| Lung total / vascular volume | 0.182 / 0.0479 mL | Table S3 |
| Lung blood flow (= cardiac output) | 605.4 mL/h | Table S3 |
| Heart blood flow / total / vascular volume | 59.3 mL/h / 0.136 mL / 0.0095 mL | Table S3 |
| Kidney blood flow / total / vascular volume | 111.25 mL/h / 0.469 mL / 0.0558 mL | Table S3 |
| Spleen blood flow / total / vascular volume | 13.3 mL/h / 0.113 mL / 0.025 mL | Table S3 |
| Liver hepatic-artery flow / total / vascular volume | 16.7 mL/h / 1.72 mL / 0.266 mL | Table S3 |
| Portal organs flow / vascular volume | 132.4 mL/h / 0.0313 mL | Table S3 |
| Remainder flow / vascular volume | 272.86 mL/h / 0.886 mL | Table S3 |

## Units in the ODE system

| Quantity | Units |
|----------|-------|
| Time | h |
| Dose | ug mRNA |
| Compartment amounts (blood, vp_*, int_*, mrna_*) | ug mRNA |
| Luciferase compartments (luc_*) | LU per mg tissue protein (homogenate assay) |
| Concentrations (`cv_bl`, `cv_lu`, ...) derived as amount / vascular volume | ug mRNA / mL |
| Volumes | mL |
| Blood flows | mL/h |
| Clearances (`cl_lu`, ..., `cl_bl`) | mL/h |
| `klnp`, `krna`, `kluc`, `kliv` | 1/h |
| `slu`, `she`, ..., `sli` | LU/mg protein per ug RNA / h |

Doses are supplied to the `blood` compartment in ug.

# Mass-balance / flux check

With LNP degradation switched off (`klnp -> 0`, `krna` decoupled from
LNP, `cl_bl -> 0`), the PBPK system is closed: total LNP mass should
be conserved across blood, blood-cleared, and all tissue
vascular / intracellular compartments. This catches structural ODE
errors that would not be visible from a numerical-fit comparison.

```{r mass-balance}
mod_no_deg <- mod |> ini(lklnp = log(1e-12)) |> ini(lcl_bl = log(1e-12))

dose_ug <- 8
ev_bal <- et(amt = dose_ug, cmt = "blood", time = 0) |>
  et(seq(0, 12, by = 0.5))
sim_bal <- as.data.frame(rxSolve(mod_no_deg, events = ev_bal))

lnp_cols <- c(
  "blood", "bldeg",
  paste0("vp_", c("lu", "he", "ki", "sp", "li", "po", "re")),
  paste0("int_", c("lu", "he", "ki", "sp", "li"))
)
sim_bal$total_lnp <- rowSums(sim_bal[, lnp_cols])

balance_summary <- sim_bal |>
  filter(time %in% c(0, 1, 4, 8, 12)) |>
  transmute(
    time_h         = time,
    total_lnp_ug   = round(total_lnp, 6),
    pct_of_dose    = round(100 * total_lnp / dose_ug, 4)
  )
knitr::kable(
  balance_summary,
  caption = "Total LNP mass with klnp = 0 and cl_bl = 0. Conservation should hold to machine precision."
)
```

Mass is conserved across the simulation window when the irreversible
loss terms (`klnp`, `cl_bl`) are zeroed, confirming that the
convective vascular flows and the per-tissue redistribution terms are
balanced in the implementation.

# Bare-LNP simulation (replicates paper Figure 2)

The packaged model file already carries the bare-LNP parameter set.
Simulate a single 8 ug IV bolus over the paper's 24-hour sampling
window and reproduce Figure 2 (blood and tissue PK in red, luciferase
activity in green).

```{r simulate}
ev <- et(amt = dose_ug, cmt = "blood", time = 0) |>
  et(seq(0, 24, by = 0.05))
sim <- as.data.frame(rxSolve(mod, events = ev))
```

```{r fig2-pk, fig.cap = "Replicates Figure 2 (red panels) of Parhiz 2024: blood and tissue LNP biodistribution as percent injected dose per gram tissue (%ID/g), bare LNPs, 8 ug IV in C57BL/6 mice."}
pk_long <- sim |>
  filter(time > 0) |>
  select(time, pid_g_blood, pid_g_lung, pid_g_heart, pid_g_kidney,
         pid_g_spleen, pid_g_liver) |>
  pivot_longer(-time, names_to = "tissue", values_to = "pid_g") |>
  mutate(tissue = factor(
    sub("pid_g_", "", tissue),
    levels = c("blood", "lung", "heart", "kidney", "spleen", "liver")
  ))

ggplot(pk_long, aes(time, pmax(pid_g, 1e-3))) +
  geom_line(color = "red") +
  facet_wrap(~tissue, scales = "free_y") +
  scale_y_log10() +
  labs(
    x = "Time (h)",
    y = "%ID / g tissue",
    title = "Bare LNP biodistribution",
    caption = "Replicates Figure 2 red traces of Parhiz et al. 2024."
  )
```

```{r fig2-pd, fig.cap = "Replicates Figure 2 (green panels) of Parhiz 2024: luciferase activity in organ homogenates after a single 8 ug bare LNP IV dose."}
luc_long <- sim |>
  filter(time > 0) |>
  select(time, luc_lung, luc_heart, luc_kidney, luc_spleen, luc_liver) |>
  pivot_longer(-time, names_to = "tissue", values_to = "luc") |>
  mutate(tissue = factor(
    sub("luc_", "", tissue),
    levels = c("lung", "heart", "kidney", "spleen", "liver")
  ))

ggplot(luc_long, aes(time, pmax(luc, 1))) +
  geom_line(color = "darkgreen") +
  facet_wrap(~tissue, scales = "free_y") +
  scale_y_log10() +
  labs(
    x = "Time (h)",
    y = "Luciferase signal (LU / mg protein)",
    title = "Bare LNP luciferase expression (homogenate assay)",
    caption = "Replicates Figure 2 green traces of Parhiz et al. 2024."
  )
```

```{r profile-table}
profile <- sim |>
  filter(time %in% c(0, 0.083, 0.5, 1, 4, 8, 12, 24)) |>
  transmute(
    time_h        = round(time, 3),
    pid_g_blood   = round(pid_g_blood, 3),
    pid_g_lung    = round(pid_g_lung, 3),
    pid_g_liver   = round(pid_g_liver, 3),
    luc_liver    = round(luc_liver, 0),
    luc_lung     = round(luc_lung, 0)
  )
knitr::kable(
  profile,
  caption = paste0("Simulated typical bare-LNP profile after a single 8 ug IV dose. ",
                   "Blood, lung, and liver represent the dominant exposure tissues.")
)
```

# Tmax / peak-magnitude check vs paper Figure 2 narrative

Paper text (Results, "Bioluminescence imaging of intact organs: Bare
vs. IgG LNPs") reports peak luciferase activity in liver at 4.5 h
after injection. The simulated liver luciferase peak time should
match this report.

```{r tmax-check}
tmax_liver <- sim |>
  filter(time > 0) |>
  slice_max(luc_liver, n = 1, with_ties = FALSE) |>
  transmute(
    tmax_liver_h = round(time, 2),
    luc_liver_peak = round(luc_liver, 0)
  )
knitr::kable(
  tmax_liver,
  caption = paste0("Simulated liver luciferase Tmax for the bare-LNP / homogenate-assay parameter set. ",
                   "Paper text reports peak liver luciferase at 4.5 h after injection (BLI of intact organs).")
)
```

# PKNCA on blood LNP concentration

The blood compartment behaves like a standard plasma concentration
profile after an IV bolus. Compute Cmax, Tmax, AUC, and apparent
terminal half-life from the simulated typical-value `Cc` (ug mRNA /
mL).

```{r pknca}
nca_input <- sim |>
  filter(!is.na(Cc), time > 0, Cc > 0) |>
  mutate(id = 1L, treatment = "8 ug IV bare LNP") |>
  select(id, time, Cc, treatment)

dose_df <- data.frame(
  id        = 1L,
  time      = 0,
  amt       = dose_ug,
  treatment = "8 ug IV bare LNP"
)

conc_obj <- PKNCA::PKNCAconc(nca_input, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_df,   amt ~ time | treatment + id)

intervals <- data.frame(
  start      = 0,
  end        = 24,
  cmax       = TRUE,
  tmax       = TRUE,
  auclast    = TRUE,
  half.life  = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- PKNCA::pk.nca(nca_data)
nca_summary <- as.data.frame(nca_res$result)
knitr::kable(
  nca_summary[, c("PPTESTCD", "PPORRES")],
  caption = "Simulated NCA parameters for bare-LNP blood concentration after a single 8 ug IV bolus."
)
```

The paper does not tabulate NCA parameters for the bare-LNP blood
profile directly; comparison is against Figure 2 panel (blood) qualitatively.
Both Cmax and Tmax align with the IV bolus expectation (Tmax = 0,
Cmax = initial Dose / V_blood = 8 ug / 1.533 mL ~= 5.22 ug/mL).

# Switching to control-IgG and PECAM-targeted parameter sets

The same structural model accommodates the IgG-coated and
PECAM-targeted parameterizations reported in the paper. To reproduce
those fits, override the `ini()` entries as follows. Numerical values
come directly from paper Tables 1-3.

## Control IgG (BLI assay) parameters

```{r igg-params, eval = FALSE}
mod_igg <- mod |>
  ini(
    lcl_lu = log(0.272),
    lcl_he = log(0.0976),
    lcl_ki = log(0.836),
    lcl_sp = log(1.41),
    lcl_li = log(5.64),
    lcl_bl = log(5.32),
    lklnp  = log(1.21),
    # BLI-assay luciferase parameters (different units than homogenate)
    lslu   = log(1.31e8),
    lshe   = log(5.73e7),
    lski   = log(6.74e6),
    lssp   = log(6.74e7),
    lsli   = log(1.20e9),
    lkluc  = log(0.250),
    lkliv  = log(0.258)
  )
```

## PECAM-targeted LNP

The PECAM model has additional terms not in the bare/IgG cases:
target-mediated extravasation (`CL_PECAM`), target-binding affinity
(`K_D`), and per-organ accessible PECAM concentration (`R_or`). To
add these, extend the `model()` block with the equilibrium-binding
quadratic from paper supplement S1 ("Lung Vasculature" equation) and
add the receptor-mediated internalization compartments (paper
supplement S1 `dARI_or/dt`). The bare-LNP file collapses to that
extended model when `R_or = CL_PECAM = 0`. The estimated PECAM
parameters from paper Table 3 are:

| Parameter | Value | Units |
|-----------|-------|-------|
| `R_lu` (lung PECAM) | 16.4 | ug RNA / mL |
| `R_he` (heart PECAM) | 1.21 | ug RNA / mL |
| `R_ki` (kidney PECAM) | 2.56 | ug RNA / mL |
| `R_sp` (spleen PECAM) | 2.98 | ug RNA / mL |
| `R_li` (liver PECAM) | 15.2 | ug RNA / mL |
| `CL_PECAM` | 0.366 | mL/h |
| `K_D` | 0.00152 | ug RNA / mL |

For PECAM-targeted luciferase parameters use the
"PECAM LNPs Homogenate" column of paper Table 2; per Methods,
non-specific-pathway luciferase production for the PECAM model is
fixed to the control-IgG values, and the PECAM-specific pathway
production rates `S_RI,or` are estimated separately.

# Assumptions and deviations

* **Bare LNP parameter set only.** The packaged model file
  reproduces the bare-LNP fit (Tables 1-2 columns "Bare LNPs" and
  "Bare LNPs Homogenate"). Control-IgG and PECAM-targeted parameter
  sets are documented in this vignette but require either an `ini()`
  override (IgG case, same structural model) or an extension of
  `model()` to add the receptor-mediated pathway (PECAM case). The
  PECAM extension is straightforward: add 5 receptor-bound-LNP
  compartments, 5 receptor-pathway mRNA compartments, 5
  receptor-pathway luciferase compartments, and the equilibrium
  free-vs-total quadratic. The structural skeleton is identical to
  the bare-LNP case with the target-binding terms zeroed.
* **Homogenate-assay luciferase parameters only.** The paper reports
  two assay formats (luciferase activity in tissue homogenates and
  bioluminescence imaging of intact organs) with different
  numerical magnitudes and decay rate constants
  (`kluc` = 0.0940 vs 0.250 1/h; `kliv` = 0.247 vs 0.258 1/h). The
  packaged file uses the homogenate-assay values. To switch to BLI,
  override `lslu` through `lsli`, `lkluc`, and `lkliv` per Table 2.
* **mRNA degradation rate (`krna`) is fixed.** The paper's Methods
  section sources `kRNA = 0.114 1/h` from prior in vitro work
  (Anderson et al. 2011 *Nucleic Acids Res* 39:9329-9338, reference
  36) and holds it constant during the PBPK fit. The file encodes
  this with `fixed(log(0.114))` in `ini()`.
* **Physiological parameters are mouse-25 g specific.** All tissue
  volumes and blood flows in the model() block are hard-coded from
  paper Table S3 (BioDMET database, 25 g mouse). For scaling to
  larger mice or other species, override the volumes and flows in
  `model()` (these are not currently exposed as `ini()` parameters).
* **No IIV, no residual error.** The paper fits a single typical-value
  parameter set per assay-formulation combination using ADAPT 5's
  maximum-likelihood estimator (Methods "Software"); RSE values
  in Tables 1-3 quantify parameter precision but no inter-individual
  variability model was fit. The packaged file is therefore a
  typical-value mechanistic simulator. For estimation use, IIV blocks
  could be added to the structural parameters guided by the Table
  1-3 RSE values.
* **Compartment naming deviation from `naming-conventions.md`.** The
  PBPK structure has 24 compartments
  (`blood`, `bldeg`, `vp_<tissue>`, `int_<tissue>`, `mrna_<tissue>`,
  `luc_<tissue>`) that do not map onto the canonical
  `central` / `peripheral1` / `depot` / `effect` vocabulary;
  `checkModelConventions("Parhiz_2024_mRNALNP")` flags every PBPK
  compartment as a non-canonical name. The naming used in this file
  follows the paper's symbolic conventions (subscripts indicate
  organ) and is the most readable mapping available. No convention
  change in the rest of nlmixr2lib is implied. The same deviation
  applies to `Shah_2012_mAb_PBPK`.
* **ADAPT 5 lung CL_PECAM divisor inconsistency.** Paper supplement
  S1 equation "Lung Vasculature" divides the CL_PECAM term by
  `Vv,lu` (vascular volume), while the published ADAPT 5 control
  stream (Data S1, mmc2.zip, `LNP Model.for` XP(3)) divides the same
  term by `Vlu` (total lung volume, line 215-216). The discrepancy
  has no effect on the bare-LNP parameterization (`CL_PECAM = 0`),
  but downstream users extending the model to PECAM-targeting should
  pick a convention before re-running the fit. The PECAM Table 3
  values were generated by the ADAPT 5 form, so retain
  `CLpe / V_lu` for that organ if exact reproducibility against the
  published `CL_PECAM` point estimate is the goal.
