---
title: "Nusinersen and pNfH neurofilament trafficking in SMA (Paris 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nusinersen and pNfH neurofilament trafficking in SMA (Paris 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Paris_2023_nusinersen_qsp"))$reference`
* Description: `r rxode2::rxode(readModelDb("Paris_2023_nusinersen_qsp"))$description`
* Article: <https://doi.org/10.1002/psp4.12890> (open access, CPT Pharmacometrics Syst Pharmacol 2023;12:196-206)
* Supplement: Appendix S1 (equations, Tables S1-S6) and Appendix S2 (the authors' complete R source code), both available from the article page.

The model describes phosphorylated neurofilament heavy subunit (pNfH), a
biomarker of axonal damage, from birth to 20 years of age in children with
spinal muscular atrophy (SMA). It extends the authors' earlier adult
neurofilament-trafficking model (Paris 2022) with growing compartment
volumes, a synaptic-pruning description of CNS development, an extra
SMA-driven leak of pNfH from the peripheral nervous system (reaction `r14`),
and an inhibition of that leak by intrathecal nusinersen. The nusinersen
amount that drives the inhibition is computed by the nine-compartment
Biliouris 2018 PK model (see the separate `Biliouris_2018_nusinersen` model),
which is embedded here with its typical rate constants.

Model time is **postnatal age in days**: every simulation starts at birth,
where the pNfH initial conditions (Table S6) are defined, and nusinersen doses
are placed at the age (in days) at which they were given. Body weight and
height enter as time-varying covariates and must be solved with
`covsInterpolation = "linear"`, because the volume growth rates are read off
their movement.

# Population

The model was parameterised on pNfH measured in CSF and plasma in the Biogen
SHINE, NURTURE, EMBRACE, ENDEAR and CHERISH trials (Table S1). 98 healthy
paediatric subjects from ENDEAR re-estimated the healthy leakage and blood
clearance; 359 SMA patients (17 untreated controls plus the pre-treatment
samples of treated patients) defined a 400-member untreated virtual
population; 278 treated patients were fitted individually to estimate the
treatment parameters; and 25 patients were held out for validation in four
age and dose groups (Table S2). Ages ranged from birth to 22 years and
intrathecal doses from 1 to 12.5 mg. All fits were normalised least-squares
fits of deterministic trajectories, not nonlinear mixed-effects estimation.

```{r population}
ui <- rxode2::rxode(readModelDb("Paris_2023_nusinersen_qsp"))
str(ui$population)
```

# Source trace

Every `ini()` value carries an in-file comment pointing to its source in
`inst/modeldb/specificDrugs/Paris_2023_nusinersen_qsp.R`. The table collects
them. "Code" means the authors' Appendix S2 R scripts
(`pNfH_SMA_model.R`, `pNfH_SMA_model_parameters.R`,
`pNfH_SMA_initial_values.R`, `PK_ASO_model.R`).

| Equation / parameter | Value | Source location |
|---|---|---|
| pNfH ODEs (7 states) | Eq. S9-S15 | Appendix S1; code `NF_SMA_model()` |
| Volume formulas | Formulas S1-S7 | Appendix S1; code (`dV_*dt` finite-difference block) |
| Pruning function and its use in r1, r3, r5 | Formulas S16-S18 | Appendix S1; code `pruning_function()`, `pruning_function_der()` |
| r14 = A + B C^(-age) | Equation 1 | Main text; code `r14 <- ...` |
| Delay chain and inhibition of r14 | Eq. S20-S24 | Appendix S1; code transit block |
| `thalf_nf_stat`, `thalf_nf_mov`, `f_nf_stat` | 90 d, 22 d, 0.9 (r2 = r4 = 0.0101/day) | Code; Table S4 r2, r4 |
| `lk_leak` (r5 = r6 base) | 1.7e-6 /day | Table S4, Table S5 |
| `q_isf_csf`, `q_endo_csf` (r7, r8) | 449.28 mL/day | Table S4 r8; code |
| `q_cran_spin`, `q_spin_cran` (r9, r10) | 7.4712, 20.148 mL/day | Table S4 |
| `q_cran_blood`, `q_spin_blood` (r11, r12) | 90, 259.2 mL/day | Table S4 |
| `lkel_pnfh` (r13) | 0.0986 /day | Code (SMA untreated); Table 1 0.10 +/- 0.01 |
| `r14_a`, `r14_b`, `lr14_c` | 1.6e-7 /day, 1.8e-5 /day, 3.3 | Table 1 / Table S4 (untreated population) |
| `p1_prune` ... `p5_prune` | -541.49, 0.15976, 527.17, 0.13413, 42.224 | Code; Table S4 |
| `lktr` (kdiff) | 0.09 /day | Table 1 / Table S4 |
| `lec50` (EC50) | 550 ng | Table 1 / Table S4 |
| `bl_pnfh_*` | 1e5, 2e6, 0.018, 3.4, 1.15, 5.4, 47 ng/mL | Table S6 (SMA untreated column); code |
| `v0_*` | 338, 38.23, 50.7, 25.88, 60, 60, 267.329 mL | Table S3 (birth column); code |
| Nusinersen PK rate constants (16) | K13 ... K91, 1/h | Code `PK_ASO_model.R` (Biliouris 2018 typical values) |
| Between-subject spread (6 etas) | variances 0.00995 ... 0.582 | Table 1 mean +/- SD, moment-matched (see Assumptions) |

# Growth inputs

The authors drive volume growth with the CDC clinical growth chart
50th-percentile male weight-for-age and length/stature-for-age curves,
linearly interpolated (Appendix S2, `Growth_charts/`). The same 242 knots are
reproduced here (birth to 240 months).

```{r cdc}
# CDC 50th-percentile male weight (kg) and length/stature (cm)
wt <- c(
  3.5302, 4.0031, 4.8795, 5.6729, 6.3914, 7.0418, 7.6304, 8.1630, 8.6448,
  9.0811, 9.4765, 9.8353, 10.1615, 10.4588, 10.7306, 10.9799, 11.2096, 11.4221,
  11.6198, 11.8048, 11.9790, 12.1440, 12.3015, 12.4528, 12.6708, 12.7415, 12.8810,
  13.0184, 13.1545, 13.2899, 13.4252, 13.5609, 13.6974, 13.8351, 13.9742, 14.1150,
  14.2578, 14.4026, 14.5496, 14.6989, 14.8505, 15.0045, 15.1608, 15.3194, 15.4803,
  15.6434, 15.8087, 15.9761, 16.1455, 16.3168, 16.4899, 16.6647, 16.8411, 17.0190,
  17.1984, 17.3791, 17.5610, 17.7440, 17.9281, 18.1132, 18.2991, 18.4859, 18.6735,
  18.8618, 19.0508, 19.2404, 19.4306, 19.6214, 19.8127, 20.0046, 20.1970, 20.3900,
  20.5836, 20.7777, 20.9724, 21.1678, 21.3638, 21.5606, 21.7581, 21.9565, 22.1557,
  22.3558, 22.5570, 22.7593, 22.9627, 23.1674, 23.3734, 23.5809, 23.7898, 24.0003,
  24.2125, 24.4265, 24.6423, 24.8601, 25.0799, 25.3019, 25.5261, 25.7526, 25.9815,
  26.2128, 26.4468, 26.6834, 26.9227, 27.1649, 27.4099, 27.6580, 27.9090, 28.1632,
  28.4206, 28.6813, 28.9453, 29.2127, 29.4836, 29.7580, 30.0360, 30.3177, 30.6031,
  30.8923, 31.1853, 31.4823, 31.7831, 32.0880, 32.3969, 32.7099, 33.0270, 33.3484,
  33.6739, 34.0036, 34.3377, 34.6760, 35.0186, 35.3656, 35.7169, 36.0726, 36.4327,
  36.7970, 37.1658, 37.5388, 37.9162, 38.2978, 38.6836, 39.0736, 39.4678, 39.8660,
  40.2683, 40.6744, 41.0844, 41.4982, 41.9156, 42.3364, 42.7607, 43.1883, 43.6190,
  44.0526, 44.4890, 44.9281, 45.3696, 45.8134, 46.2592, 46.7068, 47.1561, 47.6067,
  48.0585, 48.5111, 48.9644, 49.4181, 49.8719, 50.3255, 50.7786, 51.2310, 51.6823,
  52.1323, 52.5806, 53.0270, 53.4711, 53.9126, 54.3513, 54.7868, 55.2188, 55.6470,
  56.0712, 56.4910, 56.9061, 57.3163, 57.7214, 58.1210, 58.5149, 58.9029, 59.2848,
  59.6603, 60.0293, 60.3916, 60.7470, 61.0954, 61.4366, 61.7706, 62.0972, 62.4164,
  62.7281, 63.0323, 63.3289, 63.6180, 63.8996, 64.1737, 64.4403, 64.6996, 64.9517,
  65.1965, 65.4344, 65.6654, 65.8897, 66.1075, 66.3190, 66.5244, 66.7239, 66.9178,
  67.1064, 67.2899, 67.4686, 67.6428, 67.8128, 67.9788, 68.1411, 68.3000, 68.4558,
  68.6087, 68.7589, 68.9065, 69.0518, 69.1947, 69.3353, 69.4735, 69.6093, 69.7423,
  69.8722, 69.9987, 70.1210, 70.2386, 70.3504, 70.4555, 70.5525, 70.5976
)
ht <- c(
  49.9889, 52.6960, 56.6284, 59.6090, 62.0770, 64.2169, 66.1253, 67.8602, 69.4591,
  70.9480, 72.3459, 73.6667, 74.9213, 76.1184, 77.2648, 78.3662, 79.4273, 80.4521,
  81.4438, 82.4054, 83.3394, 84.2478, 85.1327, 85.9956, 86.4522, 86.8616, 87.6525,
  88.4233, 89.1755, 89.9104, 90.6291, 91.3324, 92.0213, 92.6964, 93.3585, 94.0082,
  94.6464, 95.2736, 95.9147, 96.5473, 97.1719, 97.7890, 98.3990, 99.0025, 99.6000,
  100.1918, 100.7783, 101.3600, 101.9373, 102.5105, 103.0799, 103.6459, 104.2087, 104.7687,
  105.3262, 105.8813, 106.4343, 106.9855, 107.5350, 108.0830, 108.6296, 109.1751, 109.7196,
  110.2631, 110.8058, 111.3477, 111.8890, 112.4296, 112.9696, 113.5090, 114.0479, 114.5861,
  115.1238, 115.6609, 116.1973, 116.7329, 117.2678, 117.8018, 118.3348, 118.8668, 119.3977,
  119.9272, 120.4554, 120.9821, 121.5072, 122.0305, 122.5520, 123.0714, 123.5886, 124.1035,
  124.6160, 125.1259, 125.6331, 126.1374, 126.6388, 127.1370, 127.6320, 128.1237, 128.6119,
  129.0966, 129.5777, 130.0550, 130.5286, 130.9983, 131.4641, 131.9260, 132.3840, 132.8381,
  133.2882, 133.7345, 134.1769, 134.6155, 135.0504, 135.4818, 135.9097, 136.3343, 136.7557,
  137.1742, 137.5899, 138.0032, 138.4143, 138.8234, 139.2310, 139.6373, 140.0427, 140.4477,
  140.8527, 141.2582, 141.6646, 142.0725, 142.4824, 142.8949, 143.3107, 143.7304, 144.1545,
  144.5838, 145.0190, 145.4607, 145.9097, 146.3665, 146.8320, 147.3066, 147.7911, 148.2859,
  148.7917, 149.3088, 149.8376, 150.3784, 150.9313, 151.4964, 152.0735, 152.6624, 153.2627,
  153.8738, 154.4951, 155.1255, 155.7642, 156.4099, 157.0612, 157.7168, 158.3751, 159.0344,
  159.6931, 160.3493, 161.0015, 161.6478, 162.2865, 162.9161, 163.5350, 164.1418, 164.7352,
  165.3140, 165.8771, 166.4236, 166.9528, 167.4641, 167.9571, 168.4313, 168.8867, 169.3231,
  169.7405, 170.1393, 170.5195, 170.8815, 171.2257, 171.5525, 171.8626, 172.1563, 172.4343,
  172.6972, 172.9456, 173.1801, 173.4014, 173.6101, 173.8067, 173.9920, 174.1665, 174.3308,
  174.4854, 174.6310, 174.7680, 174.8969, 175.0182, 175.1323, 175.2398, 175.3410, 175.4362,
  175.5259, 175.6104, 175.6901, 175.7652, 175.8360, 175.9028, 175.9658, 176.0254, 176.0816,
  176.1348, 176.1850, 176.2326, 176.2776, 176.3202, 176.3606, 176.3989, 176.4352, 176.4697,
  176.5024, 176.5335, 176.5630, 176.5911, 176.6179, 176.6433, 176.6676, 176.6907, 176.7127,
  176.7337, 176.7538, 176.7730, 176.7913, 176.8088, 176.8255, 176.8415, 176.8492
)
cdc <- data.frame(
  age_mo = c(0, seq(0.5, 22.5, 1), 24, seq(24.5, 239.5, 1), 240),
  wt = wt,
  ht = ht
)
stopifnot(nrow(cdc) == length(wt), nrow(cdc) == length(ht))

cdc_wt <- function(age_yr) approx(cdc$age_mo / 12, cdc$wt, age_yr, rule = 2)$y
cdc_ht <- function(age_yr) approx(cdc$age_mo / 12, cdc$ht, age_yr, rule = 2)$y

# One subject from birth: observation rows on `grid` (days), nusinersen
# doses (mg) at `dose_day`, and CDC-median covariates on every row. `lead`
# shifts the covariate lookup forward in age (in years).
make_events <- function(grid, dose_day = numeric(0), dose_mg = numeric(0),
                        id = 1L, lead = 0) {
  obs <- data.frame(time = grid, evid = 0L, amt = 0, cmt = "pnfh_blood")
  dos <- data.frame(time = dose_day, evid = rep(1L, length(dose_day)),
                    amt = dose_mg, cmt = rep("csf", length(dose_day)))
  bind_rows(obs, dos) |>
    mutate(id = id,
           WT = cdc_wt(time / 365 + lead),
           HT = cdc_ht(time / 365 + lead)) |>
    arrange(time, desc(evid)) |>
    select(id, time, evid, amt, cmt, WT, HT)
}

# The model has no residual-error block, so rxSolve() warns "No sigma
# parameters", and a zeroRe() model notes that its etas are zero; muffle
# only those two conditions.
solve_model <- function(model, events, ...) {
  withCallingHandlers(
    rxode2::rxSolve(model, events, covsInterpolation = "linear",
                    returnType = "data.frame", ...),
    warning = function(w) {
      if (grepl("No sigma", conditionMessage(w))) invokeRestart("muffleWarning")
    },
    message = function(m) {
      if (grepl("treated as zero", conditionMessage(m))) invokeRestart("muffleMessage")
    }
  )
}

mod_typ <- rxode2::zeroRe(ui)

# Healthy paediatric scenario: no SMA leak, healthy blood clearance and
# healthy birth concentrations (Table S6 healthy column, code
# 'healthy_pediatric').
healthy <- c(r14_a = 0, r14_b = 0, lkel_pnfh = log(0.33),
             bl_pnfh_endo = 0.05, bl_pnfh_csf_cranial = 0.1,
             bl_pnfh_csf_spinal = 0.18, bl_pnfh_blood = 0.8)
```

# Agreement with the authors' deposited code

The authors' R code was run with deSolve (lsodes) for its three built-in
scenarios: healthy, average SMA untreated, and its example treatment (12 mg
at ages 1 year + 0, 14, 77, 250, 375, 500, 625, 750, 875 and 1000 days, with
`EC50` = 550, the Table 1 estimate). The pNfH concentrations it produced at
selected ages are tabulated below and compared with this model. The code
evaluates body weight and height one day ahead of the current age (its
finite-difference step), so the comparison feeds the covariates with the same
one-day lead, and uses the code's healthy blood value at birth (8 ng/mL; see
Assumptions).

```{r identity}
ref_untx <- data.frame(
  day = c(91, 182, 365, 730, 1095, 1825, 3650, 5475, 7300),
  h_csf = c(99.5, 123.0, 173.7, 265.1, 340.2, 458.5, 613.5, 675.0, 622.4),
  h_blood = c(250.7, 257.7, 286.6, 331.3, 370.3, 384.5, 285.6, 175.3, 141.8),
  u_csf = c(4342.7, 3994.1, 3197.2, 1754.9, 1037.8, 751.8, 920.2, 1014.5, 936.4),
  u_blood = c(26081.9, 18428.7, 11267.7, 4960.5, 2809.7, 1791.3, 1260.0, 784.8, 637.5)
)
ref_tx <- data.frame(
  day = 365 + c(14, 30, 60, 100, 365, 1000),
  csf = c(240.5, 188.4, 191.4, 201.4, 265.9, 389.1),
  blood = c(5513.6, 1930.9, 1034.4, 1004.0, 1104.4, 1281.0)
)

ev_id <- make_events(0:7300, lead = 1 / 365)
sim_h <- solve_model(mod_typ, ev_id, params = c(healthy, bl_pnfh_blood = 8))
sim_u <- solve_model(mod_typ, ev_id)

dose_day_ex <- 365 + c(0, 14, 77, 250, 375, 500, 625, 750, 875, 1000)
ev_tx <- make_events(0:1825, dose_day_ex, rep(12, 10), lead = 1 / 365)
sim_tx <- solve_model(mod_typ, ev_tx)

at <- function(sim, day, col) sim[[col]][match(day, sim$time)]
cmp_untx <- ref_untx |>
  mutate(
    h_csf_sim = at(sim_h, day, "pnfh_csf_pgml"),
    h_blood_sim = at(sim_h, day, "pnfh_blood_pgml"),
    u_csf_sim = at(sim_u, day, "pnfh_csf_pgml"),
    u_blood_sim = at(sim_u, day, "pnfh_blood_pgml")
  )
cmp_tx <- ref_tx |>
  mutate(csf_sim = at(sim_tx, day, "pnfh_csf_pgml"),
         blood_sim = at(sim_tx, day, "pnfh_blood_pgml"))

cmp_untx |>
  mutate(age_years = round(day / 365, 2)) |>
  select(age_years, h_csf, h_csf_sim, h_blood, h_blood_sim,
         u_csf, u_csf_sim, u_blood, u_blood_sim) |>
  mutate(across(-age_years, \(x) signif(x, 4))) |>
  rename("Age (years)" = age_years,
         "Healthy CSF, code" = h_csf, "Healthy CSF, model" = h_csf_sim,
         "Healthy blood, code" = h_blood, "Healthy blood, model" = h_blood_sim,
         "SMA CSF, code" = u_csf, "SMA CSF, model" = u_csf_sim,
         "SMA blood, code" = u_blood, "SMA blood, model" = u_blood_sim) |>
  knitr::kable(caption = "pNfH (pg/mL) in spinal CSF and blood, healthy and SMA untreated.")

cmp_tx |>
  mutate(days_after_first_dose = day - 365) |>
  select(days_after_first_dose, csf, csf_sim, blood, blood_sim) |>
  mutate(across(c(csf_sim, blood_sim), \(x) signif(x, 4))) |>
  rename("Days after first dose" = days_after_first_dose,
         "CSF, code" = csf, "CSF, model" = csf_sim,
         "Blood, code" = blood, "Blood, model" = blood_sim) |>
  knitr::kable(caption = "pNfH (pg/mL) during the authors' example treatment from age 1 year.")

rel_err <- c(
  with(cmp_untx, abs(c(h_csf_sim / h_csf, h_blood_sim / h_blood,
                       u_csf_sim / u_csf, u_blood_sim / u_blood) - 1)),
  with(cmp_tx, abs(c(csf_sim / csf, blood_sim / blood) - 1))
)
# Same equations and parameters on both sides; the residual difference is
# integrator tolerance plus the rounding of the tabulated reference to one
# decimal place.
stopifnot(!anyNA(rel_err), max(rel_err) < 0.005)
```

The embedded nusinersen PK can be checked the same way: the authors'
`PK_ASO_model()` returns 549,299, 706,509, 761,354 and 743,796 ng of
nusinersen summed over the three spinal-cord segments 0.5, 1, 2 and 7 days
after a single 12 mg dose.

```{r aso-check}
sim_pk <- solve_model(mod_typ, make_events(c(0, 0.5, 1, 2, 7), 0, 12))
aso_ref <- c(549299.1, 706508.5, 761353.6, 743795.9)
aso_sim <- at(sim_pk, c(0.5, 1, 2, 7), "aso_sc")
data.frame(day = c(0.5, 1, 2, 7), code = aso_ref, model = signif(aso_sim, 7)) |>
  rename("Days after dose" = day, "Spinal-cord nusinersen, code (ng)" = code,
         "Spinal-cord nusinersen, model (ng)" = model) |>
  knitr::kable()
stopifnot(max(abs(aso_sim / aso_ref - 1)) < 0.001)
```

# Volume growth (Figure S1, Table S3)

```{r volumes}
grid <- seq(0, 7300, 7)
sim_u_w <- solve_model(mod_typ, make_events(grid))
sim_h_w <- solve_model(mod_typ, make_events(grid), params = healthy)

vol_long <- sim_u_w |>
  select(time, starts_with("vol_")) |>
  pivot_longer(-time, names_to = "compartment", values_to = "volume")
ggplot(vol_long, aes(time / 365, volume)) +
  geom_line() +
  facet_wrap(~compartment, scales = "free_y") +
  labs(x = "Age (years)", y = "Volume (mL)",
       caption = "Replicates Figure S1 of Paris 2023.")

vol20 <- sim_u_w |> filter(time == max(time)) |> select(starts_with("vol_"))
tab_s3 <- data.frame(
  compartment = c("vol_cns", "vol_pns", "vol_isf", "vol_endo",
                  "vol_csf_cranial", "vol_csf_spinal", "vol_blood"),
  table_s3 = c(1413, 135, 209, 91, 75, 75, 5436)
) |>
  mutate(model = signif(unlist(vol20[compartment]), 5),
         pct_diff = round(100 * (model / table_s3 - 1), 1))
tab_s3 |>
  rename("Compartment" = compartment, "Table S3, 20 years (mL)" = table_s3,
         "Model, 20 years (mL)" = model, "Difference (%)" = pct_diff) |>
  knitr::kable()
# Table S3 prints CNS 1413 mL; the code's own CNS formula freezes at 18 years
# and gives 1395 mL, which is the value its ISF entry (209 = 0.15 x 1395)
# was computed from. All other entries agree to rounding.
stopifnot(all(abs(tab_s3$pct_diff[-1]) < 0.5), abs(tab_s3$pct_diff[1]) < 1.5)
```

# Healthy and untreated SMA trajectories (Figure 2)

```{r figure2}
fig2 <- bind_rows(
  sim_h_w |> mutate(scenario = "Healthy"),
  sim_u_w |> mutate(scenario = "SMA untreated (average)")
) |>
  select(time, scenario, "CSF spinal pNfH" = pnfh_csf_pgml,
         "Blood pNfH" = pnfh_blood_pgml) |>
  pivot_longer(c("CSF spinal pNfH", "Blood pNfH"),
               names_to = "matrix", values_to = "pnfh")
ggplot(fig2, aes(time / 365, pnfh, colour = scenario)) +
  geom_line() +
  scale_y_log10() +
  facet_wrap(~matrix, scales = "free_y") +
  labs(x = "Age (years)", y = "pNfH (pg/mL)", colour = NULL,
       caption = "Replicates the typical curves of Figure 2 of Paris 2023.") +
  theme(legend.position = "bottom")
```

As in Figure 2, the average SMA trajectory starts far above the healthy one
(47,000 vs about 800 pg/mL in blood at birth, Table S6), falls steeply over
the first two years as the age-dependent part of `r14` decays, and settles
within a few-fold of the healthy curve in later childhood.

```{r r14}
r14 <- sim_u_w |>
  transmute(age = time / 365, Total = r14, `Age-independent term (A)` = r14a,
            `Age-dependent term (B C^-age)` = r14 - r14a) |>
  pivot_longer(-age, names_to = "term", values_to = "rate")
ggplot(r14, aes(age, rate, linetype = term)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Age (years)", y = "r14 (1/day)", linetype = NULL,
       caption = "Replicates Figure S2 of Paris 2023.") +
  theme(legend.position = "bottom")
```

# Treatment: delay and potency

The four-stage delay chain has a mean transit time of `4 / kdiff`:

```{r delay}
kdiff <- exp(ui$theta[["lktr"]])
c(kdiff_per_day = kdiff, mean_transit_days = 4 / kdiff)
stopifnot(abs(4 / kdiff - 44.4) < 0.1)
```

The paper describes this as "an average time delay of approximately 40 days"
between the input and the effective drug signal.

EC50 is printed as 5.5 x 10^2 in both Table 1 and Table S4, while the
authors' example script uses 5.5 x 10^3. Because the spinal-cord nusinersen
amount after a 12 mg dose (about 7 x 10^5 ng) is three orders of magnitude
above either value, the choice matters only in the first two to three weeks
after the first dose:

```{r ec50}
sim_tx_hi <- solve_model(mod_typ, make_events(0:1825, dose_day_ex, rep(12, 10)),
                         params = c(lec50 = log(5500)))
sim_tx_lo <- solve_model(mod_typ, make_events(0:1825, dose_day_ex, rep(12, 10)))
ec50_cmp <- bind_rows(
  sim_u |> filter(time <= 1825) |> mutate(scenario = "Untreated"),
  sim_tx_lo |> mutate(scenario = "Treated, EC50 = 550"),
  sim_tx_hi |> mutate(scenario = "Treated, EC50 = 5500")
)
ggplot(filter(ec50_cmp, time >= 300), aes(time / 365, pnfh_blood_pgml, colour = scenario)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Age (years)", y = "Blood pNfH (pg/mL)", colour = NULL,
       caption = "12 mg loading and maintenance from age 1 year (the authors' example).") +
  theme(legend.position = "bottom")
```

# Validation groups (Table S2, Figure 5)

Table S2 gives the median dosing schedule and the mean age at the first dose
for the four held-out validation groups. Each is simulated here from birth,
untreated and treated, as the typical subject and as a cohort of 100 virtual
patients drawing the between-subject spread from Table 1. The paper
additionally rescaled its 300 sampled curves so their mean matched each
group's mean measured pNfH before treatment, a step that depends on the
measured data and is not reproduced; the bands below are therefore the
model's prior spread for a subject of that age, not the paper's refined
bands.

```{r groups}
groups <- list(
  list(name = "Group 1 (age 0.61 y, plasma)", age = 0.61, end = 2.5,
       day = c(0, 14, 28, 63), mg = c(10.8, 10.8, 10.8, 10.8)),
  list(name = "Group 2 (age 0.36 y, plasma)", age = 0.36, end = 2.5,
       day = c(0, 14, 28), mg = c(10.3, 10.3, 11.8)),
  list(name = "Group 3 (age 0.31 y, CSF + plasma)", age = 0.31, end = 5,
       day = c(0, 14, 28, 56, 184, 297, 428, 590, 703, 836, 954),
       mg = c(10.3, 10.3, 10.3, 10.3, 10.8, 11.3, 11.3, 11.3, 12, 12, 12)),
  list(name = "Group 4 (age 1.4 y, CSF + plasma)", age = 1.4, end = 7,
       day = c(0, 14, 28, 64, 182, 302, 1031, 1155, 1268),
       mg = c(11.3, 11.3, 11.3, 11.3, 11.3, 12, 12, 12, 12))
)

n_cohort <- 100
rxode2::rxSetSeed(20230215)
grp_typ <- list()
grp_vpc <- list()
for (g in groups) {
  grid_g <- seq(0, round(g$end * 365), 7)
  dose_g <- round(g$age * 365) + g$day
  ev_untx <- make_events(grid_g)
  ev_tx_g <- make_events(grid_g, dose_g, g$mg)
  grp_typ[[g$name]] <- bind_rows(
    solve_model(mod_typ, ev_untx) |> mutate(arm = "Untreated"),
    solve_model(mod_typ, ev_tx_g) |> mutate(arm = "Treated")
  ) |>
    mutate(group = g$name, start = g$age)
  ev_tx_n <- bind_rows(lapply(seq_len(n_cohort), \(i) make_events(grid_g, dose_g, g$mg, id = i)))
  grp_vpc[[g$name]] <- solve_model(ui, ev_tx_n) |>
    group_by(time) |>
    summarise(lo = quantile(pnfh_blood_pgml, 0.025),
              med = median(pnfh_blood_pgml),
              hi = quantile(pnfh_blood_pgml, 0.975), .groups = "drop") |>
    mutate(group = g$name, start = g$age)
}
grp_typ <- bind_rows(grp_typ)
grp_vpc <- bind_rows(grp_vpc)

ggplot() +
  geom_ribbon(data = filter(grp_vpc, time / 365 > start - 0.3), aes(time / 365, ymin = lo, ymax = hi),
              fill = "steelblue", alpha = 0.25) +
  geom_line(data = filter(grp_typ, time / 365 > start - 0.3),
            aes(time / 365, pnfh_blood_pgml, colour = arm)) +
  facet_wrap(~group, scales = "free") +
  scale_colour_manual(values = c(Treated = "blue", Untreated = "orange")) +
  labs(x = "Age (years)", y = "Blood pNfH (pg/mL)", colour = "Typical subject",
       caption = paste("Shaded: 2.5th-97.5th percentile of 100 treated virtual patients.",
                       "Compare Figure 5 (a, b, d, f) of Paris 2023.")) +
  theme(legend.position = "bottom")

ggplot(filter(grp_typ, grepl("Group [34]", group), time / 365 > start - 0.3),
       aes(time / 365, pnfh_csf_pgml, colour = arm)) +
  geom_line() +
  facet_wrap(~group, scales = "free") +
  scale_colour_manual(values = c(Treated = "blue", Untreated = "orange")) +
  labs(x = "Age (years)", y = "Spinal CSF pNfH (pg/mL)", colour = "Typical subject",
       caption = "Compare Figure 5 (c, e) of Paris 2023.") +
  theme(legend.position = "bottom")
```

Figure 5 shows blood pNfH falling within roughly two months of the first
dose from the 10,000-20,000 pg/mL range to near 1,000 pg/mL, with the
untreated mean declining only slowly. The typical treated subject reproduces
that drop:

```{r groups-check}
grp_ratio <- grp_typ |>
  mutate(t_after = time - round(start * 365)) |>
  filter(t_after >= 84, t_after < 91) |>
  group_by(group) |>
  summarise(
    blood_untreated = pnfh_blood_pgml[arm == "Untreated"][1],
    blood_treated = pnfh_blood_pgml[arm == "Treated"][1],
    .groups = "drop"
  ) |>
  mutate(ratio = blood_treated / blood_untreated)
grp_ratio |>
  mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
  rename("Group" = group, "Untreated (pg/mL)" = blood_untreated,
         "Treated (pg/mL)" = blood_treated, "Treated / untreated" = ratio) |>
  knitr::kable(caption = "Typical blood pNfH about 12 weeks after the first dose.")
# Deterministic typical-value solve: twelve weeks in, the treated curve is
# 5-16% of the untreated one; a mis-scaled dose, EC50 or delay rate moves
# that by far more than the margin left here.
stopifnot(nrow(grp_ratio) == 4, all(grp_ratio$ratio < 0.25))
```

# Non-compartmental analysis

There is no NCA comparison: the paper reports no exposure metrics, and the
model's outputs are endogenous biomarker concentrations that start at a
non-zero birth value rather than drug concentrations after a dose. The
embedded nusinersen PK is validated against the authors' code above, and the
whole system is validated against the trajectories the authors' code
produces.

# Assumptions and deviations

* **The deposited code is followed where it disagrees with the printed
  tables.** Appendix S2 is described by the authors as the complete source
  code of the model; it reproduces Figure 2, and its values are used here.
  The disagreements are:
  - Pruning rates: the code has P2 = 0.1597563 and P4 = 0.1341329; Table S4
    prints 0.161 and 0.131. P1, P3 and P5 in Table S4 are exact roundings of
    the code values, P2 and P4 are not. With the Table S4 values every pNfH
    concentration (CNS, spinal CSF, blood) is up to about 8% higher.
  - r7: Table S4 prints 0.33; the code uses 449.28 mL/day, and Table S4
    itself gives r8 = 449.28 mL/day "assumed equal to r7".
  - r1 and r3: Table S4 prints 12300.21 ng/(mL day); the code (and the
    Table S4 description "at equilibrium with degradation for the initial
    concentration") sets them to r2 times the birth concentration,
    about 1008 ng/(mL day) for the CNS. The code also applies the pruning
    shape to r3, where Formula S17 names only r1.
  - CSF volumes: Formulas S5 and S6 are printed as steps (60, 65, 67.7, 70,
    75 mL); the code interpolates linearly through 60, 65, 67.5, 70 and 75 mL
    at 0, 0.375, 0.75, 1.5 and 2 years. The linear form is used because a
    step would put an instantaneous dilution in the concentration equations.
  - CNS volume at 20 years: Table S3 prints 1413 mL; the code gives 1395 mL
    (see the Table S3 comparison above).
  - Healthy blood pNfH at birth: Table S6 gives 800 pg/mL, the code 8 ng/mL
    (8000 pg/mL). With a blood half-life of about 2 days the two converge
    within the first weeks; the Table S6 value is used for the healthy
    scenario except in the code comparison.
* **Volumes at birth do not equal the growth formulas at birth.** The birth
  volumes (Table S3) are the code's initial conditions; the formulas
  evaluated at the CDC birth weight give a larger CNS (424 vs 338 mL) and
  blood (282 vs 267 mL) volume, and the volume states relax onto them within
  hours. That relaxation dilutes the CNS pNfH by about 23% on the first day
  of life. This is the authors' implementation and is kept.
* **Volume dynamics.** The code relaxes each volume onto its target at
  365 per day (its finite-difference step of one day, expressed in years),
  with the target evaluated one day ahead in age. Here the CSF and blood
  formula switches use the same one-day lead, but body weight and height are
  read at the current time, so supply them at the subject's age. The neural
  and CSF volumes stop growing at 18 years and the blood volume at 20.
* **EC50 units.** Tables 1 and S4 give EC50 in ng/mL, but the driver of the
  delay chain is the summed nusinersen *amount* (ng) in the three
  spinal-cord compartments of the PK model (Appendix S1, Eq. S20; code
  `PK_ASO_model()`), so EC50 is an amount. The Table 1 value 550 is used; the
  code's example script uses 5500 (compared above).
* **Embedded nusinersen PK.** The code uses the Biliouris 2018 typical rate
  constants without the body-weight scaling and between-subject variability
  of the published PK model, although the main text says the PK model was
  translated to children allometrically. The code's version is used.
  The code's dosing helper adds the first dose's amount at every dosing
  time; this model doses the amount given, which only matters for schedules
  with unequal doses such as those of Table S2.
* **Default parameter set.** The defaults are the code's average SMA
  untreated scenario (r13 and r14 from the untreated virtual population,
  Table 1 first column) with the treatment parameters from the individual
  treatment fits (Table 1 second column), as in the authors' example script.
  The treatment-fit means for r13 and r14 (0.12 /day; 1.04e-7 /day,
  1.56e-5 /day, 3.25) can be supplied through `params`.
* **Between-subject spread.** The paper reports parameter distributions as
  mean +/- SD of the 400-member untreated virtual population and of the
  individual treatment fits (Table 1), not as estimated random effects.
  They are encoded as independent log-normal etas whose variance
  `log(1 + (SD / mean)^2)` matches each CV (r13 10%, r14A 62.5%, r14B 33%,
  r14C 27%, EC50 38%, kdiff 89%). The shape of the distributions and their
  correlations are not reported. There is no residual-error model; the
  paper fitted with normalised least squares and reports no residual
  variance.
* **Growth curves.** The CDC 50th-percentile male curves are the authors'
  default for every subject; sex-specific or individual curves can be
  supplied as `WT` and `HT`.
* **Newborns.** The paper notes the model underpredicts CSF pNfH in the first
  three months of life (NURTURE newborns were excluded from the treatment
  training set).
