---
title: "Efineptakin alfa / rhIL-7-hyFc (Park 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Efineptakin alfa / rhIL-7-hyFc (Park 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
rxode2::rxSetSeed(20250101)
set.seed(20250101)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Park_2025_efineptakin_alfa"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.2147/DDDT.S564085>
* Open-access full text: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC12539412/>

rhIL-7-hyFc (efineptakin alfa) is a hybrid-Fc-fused long-acting recombinant
human interleukin-7. Park 2025 fitted a sequential population PK then PK-PD
model to a phase 1b dose-escalation study (NCT03478995) in patients with
advanced solid tumours, and used it to propose a recommended phase 2 dose.

# Population

```{r population, echo = FALSE}
pop <- ui$population
tibble::tibble(
  Field = names(pop),
  Value = vapply(pop, function(x) paste(
    if (is.null(names(x))) as.character(x) else paste0(names(x), " ", x),
    collapse = "; "), character(1))
) |>
  knitr::kable(caption = "Population metadata carried by the model file.")
```

Thirty-five adults with locally advanced, recurrent or metastatic solid tumours
received intramuscular rhIL-7-hyFc: 29 in an 8-level dose escalation from 0.06
to 1.7 mg/kg every 3 weeks, and 6 at 1.2 mg/kg every 6 weeks. The analysis data
set holds 402 serum concentrations and 256 absolute lymphocyte counts (ALC)
across 75 treatment cycles (Park 2025 Results, "Patient Data"). Demographics are
Table 1: mean age 57.9 years, 19 male / 16 female, all Asian, mean body weight
67.1 kg in males and 55.3 kg in females. The pooled cohort mean weight used
throughout this vignette is therefore
`r signif((19 * 67.1 + 16 * 55.3) / 35, 3)` kg.

Baseline ALC was a median of 1268 cells/uL (range 341-2453) with a Table 1 mean
of 1314.4 cells/uL, and baseline endogenous serum IL-7 a median of 0.05 ng/mL
(range 0.02-0.23). The ELISA cannot separate endogenous interleukin-7 from
drug, so that baseline is carried in the observation model rather than
subtracted from the data.

# Source trace

Every `ini()` entry in `inst/modeldb/specificDrugs/Park_2025_efineptakin_alfa.R`
carries an in-file comment naming its source location. They are collected here
for review.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lka` | 0.341 1/h | Table 2, `Ka` |
| `logitfdepot` | 0.346 | Table 2, `BIOF` |
| `lfdepot` | 1 (fixed) | Table 2, `BIOA`; footnote "fixed at 1" |
| `ltlag` | 22.5 h | Table 2, `ALAG2` |
| `lvc` | 1290.0 L | Table 2, `V_C` |
| `lvp` | 15000.0 L | Table 2, `V_P` |
| `lq` | 10.80 L/h | Table 2, `Q` |
| `ltclchange` | 350.0 h | Table 2, `T_CLchange` |
| `lcl` | 12.40 L/h | Table 2, `CL_M,<TCLchange` |
| `e_sexf_cl` | log(4.82 / 12.40) | Table 2, `CL_F,<TCLchange` = 4.82 vs male 12.40 |
| `lcl_late` | 45.80 L/h | Table 2, `CL_>TCLchange` |
| `lbl_il7` | 0.05 ng/mL | Results, "median baseline serum IL-7 concentration was 0.05 ng/mL" |
| `etalka` / `etalfdepot` / `etalvc` / `etalq` / `etalcl` | 0.609 / 0.377 / 0.198 / 1.81 / 0.613 | Table 2 IIV rows (variances; see below) |
| `etalbl_il7` | log(0.275^2 + 1) | Page 9309 observation equation (var(omega_2) = var(epsilon_2)) + Table 2 `epsilon` |
| `propSd_Cil7` | 0.275 | Table 2, `epsilon (proportional)`, footnote "represented as CV" |
| `lmtt` | 146.0 h | Table 3, `MTT` |
| `lgamma` | 0.130 | Table 3, `gamma` |
| `lkout` | 0.051 1/h | Table 3, `Kcirc` |
| `lemax` | 0.155 | Table 3, `Emax` |
| `lec50` | 0.066 ng/mL | Table 3, `EC50` |
| `lrbase_alc` | 1314.4 cells/uL | Table 1, ALC mean (not an estimated parameter -- see Errata) |
| `etalmtt` / `etalemax` / `etalec50` | 0.127 / 3.900 / 0.11 | Table 3 IIV rows |
| `addSd_ALC` / `propSd_ALC` | 2.450 / 0.195 | Table 3, `epsilon_1` (SD) and `epsilon_2` (CV) |
| Two-depot absorption, `F1` / `F2` logistic split | n/a | Table 2 footnote (verbatim logistic formula) |
| Observation equation `C_obs = C_pred + C_BSL + errors` | n/a | Page 9309 |
| Step clearance at `TCLchange` (NONMEM MTIME) | n/a | Page 9309, "Final Pharmacokinetic Model" |
| Lymphopoiesis ODEs (`Prol`, 3 transits, `Circ`) | n/a | Page 9310 display equations + Figure 2 legend |
| `E = Emax * CONC / (EC50 + CONC)` | n/a | Page 9310 |
| `KProl = Ktr` | derived | Forced by `dProl/dt = 0` at baseline (see Errata) |

## The IIV column is on the variance scale

Tables 2 and 3 print each random effect as `omega (CV %)`. The table footnote
gives `CV% = 100 * sqrt(exp(omega) - 1)`, which only reproduces the printed
percentages if the tabulated number is a **variance**. This is checked
mechanically rather than assumed:

```{r iiv-scale}
iiv <- tibble::tribble(
  ~parameter,   ~omega2, ~published_cv,
  "Ka",           0.609,  91.6,
  "BIOA",         0.377,  67.7,
  "Vc",           0.198,  46.8,
  "Q",            1.81,  226.1,
  "CL (IIV+IOV)", 0.613,  91.9,
  "MTT",          0.127,  36.8,
  "Emax",         3.900, 695.7
) |>
  mutate(
    cv_if_variance = 100 * sqrt(exp(omega2) - 1),
    cv_if_sd       = 100 * sqrt(exp(omega2^2) - 1),
    abs_err_var    = abs(cv_if_variance - published_cv)
  )

iiv |>
  select(parameter, omega2, published_cv, cv_if_variance, cv_if_sd) |>
  rename("Parameter" = parameter, "Tabulated value" = omega2,
         "Published CV%" = published_cv, "CV% if variance" = cv_if_variance,
         "CV% if SD" = cv_if_sd) |>
  knitr::kable(digits = 2, caption = "Reading the tabulated omega column as a variance reproduces every published CV%.")

# Every row reproduces to better than 0.1 CV points under the variance reading.
stopifnot(all(iiv$abs_err_var < 0.1))
```

`IIV_EC50` is excluded from the assertion above only because the table rounds it
to two decimals: `0.11` prints as 34.1% while the paper prints 34.7%, which is
recovered exactly from an unrounded 0.1137. The reading is the same.

# Structural checks

The paper's display equations on page 9310 contain three defects. Each is
corrected in the model file, and each correction is *forced* by the paper's own
numbers rather than chosen. The checks below re-derive them from the packaged
model, so a regression in the model file fails the render.

```{r load-model}
mod <- readModelDb("Park_2025_efineptakin_alfa")
mp  <- ui$theta

EMAX  <- exp(mp[["lemax"]])
GAMMA <- exp(mp[["lgamma"]])
CIRC0 <- exp(mp[["lrbase_alc"]])
KOUT  <- exp(mp[["lkout"]])
KTR   <- 4 / exp(mp[["lmtt"]])
WT    <- (19 * 67.1 + 16 * 55.3) / 35   # pooled cohort mean, Table 1
c(Emax = EMAX, gamma = GAMMA, Circ0 = CIRC0, kout = KOUT, ktr = KTR, WT = WT)
```

## 1. The feedback term is printed upside down

Page 9310 typesets

`dProl/dt = KProl * Prol * (1 + E) * (Circ / Circ0)^gamma - Ktr * Prol`

with `Circ` in the *numerator*. Setting `dProl/dt = 0` with `KProl = Ktr` gives
a maximum attainable ALC of `(1 + Emax)^(-1/gamma)` relative to baseline. That
is a **fall**, which contradicts the paper's own results throughout ("a
dose-dependent increase in ALC"; "approximately 1.8- to 2.3-fold increase from
baseline"; "+1088 to +1802 cells/uL"). Inverting the ratio to `(Circ0 / Circ)^gamma`
-- the Friberg 2002 form, the form used by every other Friberg model in this
library, and the mechanism the Discussion describes -- gives a rise.

```{r feedback-direction}
as_printed <- (1 + EMAX)^(-1 / GAMMA)   # Circ in the numerator
corrected  <- (1 + EMAX)^( 1 / GAMMA)   # Circ0 in the numerator (implemented)
c(`as printed` = as_printed, `as implemented` = corrected)

# The paper reports 1.8-2.3x rises, so only one direction is even feasible.
stopifnot(as_printed < 1, corrected > 2.3)
```

## 2. The ceiling is reproduced by simulation

The corrected form predicts an exact asymptote: under a concentration clamped
far above `EC50`, ALC must converge to `Circ0 * (1 + Emax)^(1/gamma)`. A long
zero-order infusion drives the system there.

```{r clamp-asymptote}
T_END <- 20000
clamp_ev <- bind_rows(
  tibble::tibble(id = 1L, SEXF = 0, time = 0, evid = 1L, amt = 5e5,
                 cmt = "central", dvid = NA_integer_, rate = 5e5 / T_END),
  tibble::tibble(id = 1L, SEXF = 0, time = seq(0, T_END, by = 50), evid = 0L,
                 amt = NA_real_, cmt = "central", dvid = 1L, rate = NA_real_)
) |>
  arrange(time, desc(evid))

clamp <- rxode2::rxSolve(rxode2::zeroRe(mod), clamp_ev,
                         returnType = "data.frame", useLinCmt = FALSE)

clamp_ratio <- tail(clamp$ALC, 1) / CIRC0
c(Cc_end_ng_mL = tail(clamp$Cc, 1), simulated_ratio = clamp_ratio,
  analytic_ratio = corrected)

# Pure numerical agreement between a solve and its own closed form: tight.
stopifnot(abs(clamp_ratio - corrected) / corrected < 1e-3)
```

## 3. `Kcirc` is a real parameter, and `KProl = Ktr` holds the system at baseline

The typeset `dCirc/dt` reads `Ktr * Transit3 - Ktr * Circ`. Read literally,
`Kcirc` -- an estimated parameter with a point estimate, a bootstrap median and
a 95% CI -- would appear nowhere in the model. It is numerically distinct from
`Ktr`, which settles it.

`KProl` has no row in Table 3 at all. Requiring the undosed system to rest at
its own stated baseline forces `KProl = Ktr` identically, and forces the initial
conditions `precursor_i(0) = Circ0 * Kcirc / Ktr`. Both are derivations, not
assumptions, and an undosed simulation must therefore hold perfectly flat.

```{r steady-hold}
c(ktr = KTR, kcirc = KOUT, ratio = KOUT / KTR)
stopifnot(abs(KOUT - KTR) / KTR > 0.5)   # not the same parameter

hold_ev <- tibble::tibble(id = 1:2, SEXF = c(0, 1)) |>
  crossing(time = seq(0, 6000, by = 24)) |>
  mutate(evid = 0L, amt = NA_real_, cmt = "central", dvid = 1L)

hold <- rxode2::rxSolve(rxode2::zeroRe(mod), hold_ev,
                        returnType = "data.frame", useLinCmt = FALSE)

hold_drift <- max(abs(hold$ALC - CIRC0)) / CIRC0
c(baseline = CIRC0, max_relative_drift = hold_drift)

# Exact by construction; the tolerance is integrator noise, not model behaviour.
stopifnot(hold_drift < 1e-8)
```

The chain holds at `r signif(CIRC0, 5)` cells/uL to a relative drift of
`r signif(hold_drift, 2)` over 250 days. This single check confirms the
`KProl = Ktr` derivation, the `Circ0 * Kcirc / Ktr` initial conditions, and the
`(Circ0 / Circ)^gamma` feedback normalisation simultaneously.

## 4. EC90 is exactly nine times EC50

The Discussion reports "The low EC90 value (0.594 ng/mL), approximately 9 times
the EC50". For a simple Emax model that ratio is an identity, so it is a clean
check that the packaged `EC50` is the published one.

```{r ec90}
EC50 <- exp(mp[["lec50"]])
ec90 <- 9 * EC50
c(EC50 = EC50, EC90 = ec90, published_EC90 = 0.594)
stopifnot(abs(ec90 - 0.594) < 5e-4)
```

## 5. The default solver path agrees with the explicit-ODE path

`rxSolve()` defaults to `useLinCmt = TRUE`, which rewrites a recognisably linear
system into a closed form. On a plain two-compartment model whose peripheral
transfer is written as the micro-constants `k12` / `k21`, that rewrite is known
to drop `peripheral1` and silently solve a *one*-compartment model -- and total
AUC is unaffected, so the obvious `AUC == Dose/CL` check does not catch it.

This model writes `k12` / `k21` inside `model()`, so the risk has to be
excluded rather than assumed. It stores `lq` / `lvp` in `ini()` (never `lk12` /
`lk21`), and the two parallel depots, the step clearance and the lymphopoiesis
chain all make the system non-representable as `linCmt()`, so the rewrite should
be skipped entirely. That is asserted rather than argued, because the failure
would be silent and would hit any downstream user calling `rxSolve()` without
the flag.

```{r lincmt-equivalence}
# Self-contained: this section runs before the cohort helper is defined.
probe_dose <- tibble::tibble(id = 1L, SEXF = 0, time = c(0, 2016), evid = 1L,
                             amt = 1.2 * WT, dvid = NA_integer_)
probe_ev <- bind_rows(
  mutate(probe_dose, cmt = "depot"),
  mutate(probe_dose, cmt = "depot2"),
  tibble::tibble(id = 1L, SEXF = 0,
                 time = sort(unique(c(seq(0, 4032, by = 12),
                                      c(0, 0.5, 1, 2, 4, 8, 24, 48) + 2016))),
                 evid = 0L, amt = NA_real_, cmt = "central", dvid = 1L)
) |>
  arrange(time, desc(evid))

probe_off <- rxode2::rxSolve(rxode2::zeroRe(mod), probe_ev,
                             returnType = "data.frame", useLinCmt = FALSE)
probe_on  <- rxode2::rxSolve(rxode2::zeroRe(mod), probe_ev,
                             returnType = "data.frame", useLinCmt = TRUE)

stopifnot(nrow(probe_on) == nrow(probe_off))
cc_gap  <- max(abs(probe_on$Cc  - probe_off$Cc))  / max(probe_off$Cc)
alc_gap <- max(abs(probe_on$ALC - probe_off$ALC)) / max(probe_off$ALC)
c(peripheral1_present = "peripheral1" %in% names(probe_on),
  max_rel_gap_Cc = cc_gap, max_rel_gap_ALC = alc_gap)

# Same drawn parameters on both sides, so any difference is pure numerical
# error and a tight bound is correct. A dropped peripheral compartment would
# move Cc by far more than this.
stopifnot("peripheral1" %in% names(probe_on), cc_gap < 1e-10, alc_gap < 1e-10)
```

# Virtual cohort

Original observed data are not publicly available (Park 2025 Data Sharing
Statement). The simulations below use the paper's own Monte Carlo design: the
dose levels and dosing intervals of its Figure 5 and Figure 6, at the pooled
cohort mean body weight.

**Typical-value trajectories are used for the quantitative replication.** This
is a deliberate choice forced by the model, not a shortcut -- see
"Assumptions and deviations". The published `IIV_Emax` of 3.900 combined with
`1/gamma = 7.69` makes a stochastic draw explosive, and the paper's reported
Monte Carlo means are reproduced by the typical-value trajectory.

```{r cohort}
TAU <- c(q6w = 6 * 7 * 24, q12w = 12 * 7 * 24)

# A grid that is dense right after each dose. A coarse grid does not resolve the
# absorption peak and understates AUC (and hence Cavg) by more than 15%.
dense_offsets <- c(0, 0.5, 1, 2, 4, 8, 12, 18, 24, 36, 48, 72, 96, 144, 192, 240)

make_arm <- function(dose_mgkg, tau, n_cycles, label, id, SEXF = 0) {
  dose_times <- tau * seq_len(n_cycles) - tau
  obs_times <- sort(unique(c(
    seq(0, n_cycles * tau, by = 24),
    as.vector(outer(dense_offsets, dose_times, "+"))
  )))
  obs_times <- obs_times[obs_times <= n_cycles * tau]
  dos <- tibble::tibble(id = id, SEXF = SEXF, time = dose_times, evid = 1L,
                        amt = dose_mgkg * WT, dvid = NA_integer_)
  bind_rows(
    mutate(dos, cmt = "depot"),
    mutate(dos, cmt = "depot2"),
    tibble::tibble(id = id, SEXF = SEXF, time = obs_times, evid = 0L,
                   amt = NA_real_, cmt = "central", dvid = 1L)
  ) |>
    mutate(regimen = label, dose_mgkg = dose_mgkg, tau = tau) |>
    arrange(id, time, desc(evid))
}

ARMS <- tidyr::expand_grid(
  dose_mgkg = c(0.05, 0.1, 0.2, 0.4, 0.6, 0.8, 1.0, 1.2),
  interval  = names(TAU)
) |>
  mutate(
    # unname() is load-bearing: TAU is a named vector, so TAU[interval] carries
    # names through into events$tau, and rxSolve()'s keep= rejects a named
    # column with "'names' attribute [n] must be the same length as the vector".
    tau      = unname(TAU[interval]),
    n_cycles = ifelse(interval == "q6w", 6L, 3L),
    label    = sprintf("%.2f mg/kg %s", dose_mgkg, interval),
    id       = row_number()
  )

events <- bind_rows(lapply(seq_len(nrow(ARMS)), function(i) {
  make_arm(ARMS$dose_mgkg[i], ARMS$tau[i], ARMS$n_cycles[i], ARMS$label[i], ARMS$id[i])
}))

stopifnot(
  # Disjoint ids are mandatory for multi-arm event tables: shared ids silently
  # collapse several regimens into one subject. Assert the arm count directly --
  # `!anyDuplicated(unique(x))` can never fail, because unique() removes the
  # duplicates the test is looking for.
  nrow(distinct(events, id, regimen)) == nrow(ARMS),
  nrow(distinct(events, id)) == nrow(ARMS),
  # No duplicated event row. `cmt` must be in the key: a dose legitimately
  # appears twice at the same (id, time, evid), once per depot.
  !anyDuplicated(events[, c("id", "time", "evid", "cmt")])
)
nrow(events)
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  rxode2::zeroRe(mod), events = events,
  keep = c("regimen", "dose_mgkg", "tau"),
  returnType = "data.frame", useLinCmt = FALSE
)

# Restrict to the final (steady-state) dosing interval of each arm.
ss <- sim |>
  group_by(id) |>
  mutate(cycle_start = max(tau) * (max(time) / max(tau) - 1)) |>
  filter(time >= cycle_start) |>
  ungroup()

trapz <- function(x, y) sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)

ss_summary <- ss |>
  group_by(regimen, dose_mgkg, tau) |>
  summarise(
    cavg_typical = trapz(time, Cc) / first(tau),
    alc_avg      = trapz(time, ALC) / first(tau),
    alc_max      = max(ALC),
    .groups      = "drop"
  ) |>
  mutate(
    # Park 2025 reports the MEAN across a simulated cohort. Cavg is proportional
    # to 1/CL, so for a log-normal CL the cohort mean exceeds the typical value
    # by exactly exp(omega^2 / 2).
    cavg_cohort_mean = cavg_typical * exp(0.613 / 2),
    alc_rise         = alc_avg - CIRC0
  )
```

# Replicate published figures

```{r figure-5, fig.cap = "Replicates Figure 5 of Park 2025: simulated serum concentration and ALC over time after repeated intramuscular rhIL-7-hyFc every 12 weeks."}
sim |>
  filter(dose_mgkg %in% c(0.6, 1.2), tau == TAU[["q12w"]]) |>
  select(time, regimen, `Total serum IL-7 (ng/mL)` = Cil7, `ALC (cells/uL)` = ALC) |>
  pivot_longer(-c(time, regimen), names_to = "endpoint", values_to = "value") |>
  mutate(week = time / (24 * 7)) |>
  ggplot(aes(week, value, colour = regimen)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~endpoint, scales = "free_y") +
  labs(x = "Time (weeks)", y = NULL, colour = NULL,
       title = "Figure 5 -- 0.6 and 1.2 mg/kg IM every 12 weeks",
       caption = "Replicates Figure 5 of Park 2025.") +
  theme(legend.position = "bottom")
```

```{r figure-6, fig.cap = "Replicates Figure 6 of Park 2025: predicted steady-state exposure-response for ALC."}
ss_summary |>
  mutate(interval = ifelse(tau == TAU[["q6w"]], "every 6 weeks", "every 12 weeks")) |>
  ggplot(aes(dose_mgkg, alc_avg, colour = interval)) +
  geom_hline(yintercept = 1500, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.7) +
  geom_point() +
  labs(x = "Dose (mg/kg)", y = "Steady-state average ALC (cells/uL)", colour = NULL,
       title = "Figure 6 -- dose-response at steady state",
       caption = paste("Replicates Figure 6 of Park 2025. Dashed line: the 1500 cells/uL",
                       "level the paper reports being exceeded by every regimen.")) +
  theme(legend.position = "bottom")
```

The paper describes "a steep rise in ALC ... between 0 and 0.6 mg/kg, with a
plateauing trend at doses above 0.6 mg/kg". That shape is asserted rather than
eyeballed: the marginal ALC gain per mg/kg above 0.6 must be far smaller than
below it.

```{r plateau}
plateau <- ss_summary |>
  filter(tau == TAU[["q12w"]]) |>
  arrange(dose_mgkg) |>
  mutate(slope = (alc_avg - lag(alc_avg)) / (dose_mgkg - lag(dose_mgkg)))

plateau |>
  select(dose_mgkg, alc_avg, slope) |>
  rename("Dose (mg/kg)" = dose_mgkg, "Steady-state average ALC (cells/uL)" = alc_avg,
         "Marginal gain (cells/uL per mg/kg)" = slope) |>
  knitr::kable(digits = c(2, 0, 0), caption = "Marginal ALC gain per mg/kg falls monotonically with dose.")

slope_below <- with(plateau, mean(slope[dose_mgkg <= 0.6], na.rm = TRUE))
slope_above <- with(plateau, mean(slope[dose_mgkg >  0.6], na.rm = TRUE))
c(`mean slope <= 0.6 mg/kg` = slope_below, `mean slope > 0.6 mg/kg` = slope_above,
  ratio = slope_below / slope_above,
  `last / first marginal gain` = tail(na.omit(plateau$slope), 1) / plateau$slope[2])

stopifnot(
  # The structural claim, and the one with real margin: the response saturates,
  # so every successive marginal gain is strictly smaller than the one before.
  all(diff(na.omit(plateau$slope)) < 0),
  # By the top of the range the marginal gain has fallen to under a quarter of
  # its value at the bottom (achieved: ~0.19).
  tail(na.omit(plateau$slope), 1) < 0.25 * plateau$slope[2],
  # Still rising above 0.6 mg/kg, just far more slowly -- which is what a
  # "plateauing trend" means. Achieved ratio ~2.9. These are typical-value
  # trajectories on a fixed grid, so the number is deterministic, but the bound
  # is set below the achieved value rather than at it.
  slope_above > 0,
  slope_below > 2 * slope_above
)
```

# PKNCA validation

The exposure quantity Park 2025 reports is the steady-state average
concentration, obtained by "normalizing AUC values to the respective dosing
interval". PKNCA computes that AUC over the final dosing interval of each arm.

```{r pknca}
nca_conc <- ss |>
  filter(!is.na(Cc)) |>
  group_by(id) |>
  mutate(time_in_cycle = time - min(time)) |>
  ungroup() |>
  select(id, regimen, time = time_in_cycle, Cc)

# Guarantee a time = 0 record per (id, regimen); PKNCA anchors AUC0-tau on it.
nca_conc <- bind_rows(
  nca_conc,
  nca_conc |> distinct(id, regimen) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, regimen, time, .keep_all = TRUE) |>
  arrange(id, regimen, time)

nca_dose <- ARMS |>
  transmute(id, regimen = label, time = 0, amt = dose_mgkg * WT)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(nca_conc), Cc ~ time | regimen + id)
dose_obj <- PKNCA::PKNCAdose(as.data.frame(nca_dose), amt ~ time | regimen + id)

intervals <- ARMS |>
  transmute(start = 0, end = tau, auclast = TRUE, cmax = TRUE, tmax = TRUE,
            cav = TRUE, regimen = label, id) |>
  as.data.frame()

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  select(regimen, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
head(nca_wide)
```

## Mass balance: AUC over a steady-state interval equals Dose / CL

At steady state every dosing interval begins well past `TCLchange`, so the late
clearance arm alone governs and `AUC(0-tau) = Dose / CL_late` exactly. This is a
per-arm identity between the solve and a closed form using the same parameters,
so a tight bound is correct here.

```{r mass-balance}
CL_LATE <- exp(mp[["lcl_late"]])

mb <- nca_wide |>
  left_join(ARMS |> select(regimen = label, dose_mgkg, tau), by = "regimen") |>
  mutate(
    dose_mg      = dose_mgkg * WT,
    # auclast is in ng/mL * h; Dose/CL is in mg/L, hence the 1000x.
    auc_expected = dose_mg / CL_LATE * 1000,
    pct_diff     = 100 * (auclast - auc_expected) / auc_expected
  )

mb |>
  select(regimen, auclast, auc_expected, pct_diff) |>
  rename("Regimen" = regimen, "AUC0-tau (ng*h/mL)" = auclast,
         "Dose / CL_late (ng*h/mL)" = auc_expected, "Difference (%)" = pct_diff) |>
  knitr::kable(digits = c(0, 1, 1, 2),
               caption = "Steady-state AUC against the Dose / CL_late identity, per arm.")

stopifnot(nrow(mb) == nrow(ARMS), all(!is.na(mb$pct_diff)))
# Residual difference is trapezoidal error on the absorption peak plus the small
# accumulation not yet washed out, not a structural discrepancy.
stopifnot(max(abs(mb$pct_diff)) < 5)
```

## Comparison against published values

Park 2025 reports no NCA table, but it does report simulated steady-state
average concentrations and average ALCs, which are the same quantities. Those
are the answer key.

```{r nca-compare}
published <- tibble::tribble(
  ~regimen,          ~cav,  ~alc_avg,
  "0.05 mg/kg q12w", 0.05,  NA,
  "0.60 mg/kg q12w", 0.54,  2439,
  "1.20 mg/kg q12w", 1.09,  2716,
  "0.60 mg/kg q6w",  NA,    2898,
  "1.20 mg/kg q6w",  2.37,  3165
)

cmp <- ss_summary |>
  select(regimen, cavg_cohort_mean, alc_avg) |>
  inner_join(published, by = "regimen", suffix = c("_sim", "_pub")) |>
  transmute(
    Regimen                        = regimen,
    `Cavg simulated (ng/mL)`       = cavg_cohort_mean,
    `Cavg published (ng/mL)`       = cav,
    `Cavg diff (%)`                = 100 * (cavg_cohort_mean - cav) / cav,
    `ALC simulated (cells/uL)`     = alc_avg_sim,
    `ALC published (cells/uL)`     = alc_avg_pub,
    `ALC diff (%)`                 = 100 * (alc_avg_sim - alc_avg_pub) / alc_avg_pub
  )
knitr::kable(cmp, digits = c(0, 3, 3, 1, 0, 0, 1),
             caption = "Simulated steady-state exposure and response against the values Park 2025 reports.")
```

```{r nca-assertions}
cavg_err <- abs(cmp$`Cavg diff (%)`)
cavg_err <- cavg_err[!is.na(cavg_err)]
alc_err  <- abs(cmp$`ALC diff (%)`)
alc_err  <- alc_err[!is.na(alc_err)]

# Guard against a lookup that silently matched nothing.
stopifnot(length(cavg_err) == 4L, length(alc_err) == 4L)

c(cavg_median_abs_pct = median(cavg_err), cavg_max_abs_pct = max(cavg_err),
  alc_median_abs_pct  = median(alc_err),  alc_max_abs_pct  = max(alc_err))

sim_cav <- function(reg) cmp$`Cavg simulated (ng/mL)`[cmp$Regimen == reg]

# The two exposures the paper reports at full precision are asserted directly,
# one regimen at a time, rather than through a summary statistic that lets a
# bad arm hide behind a good one. A mis-transcribed clearance, dose or unit
# would move either by tens of percent.
stopifnot(
  abs(sim_cav("0.60 mg/kg q12w") - 0.54) / 0.54 < 0.05,   # achieved ~2.6%
  abs(sim_cav("1.20 mg/kg q12w") - 1.09) / 1.09 < 0.05    # achieved ~1.7%
)

# 0.05 mg/kg q12w is NOT a 7.6% miss: the paper prints "0.05 ng/mL" to two
# decimal places, so any value in [0.045, 0.055) rounds to it. The simulated
# 0.046 is inside that interval, which is the strongest statement the published
# precision supports.
stopifnot(sim_cav("0.05 mg/kg q12w") >= 0.045, sim_cav("0.05 mg/kg q12w") < 0.055)

# 1.20 mg/kg q6w is checked against the paper's OWN q12w number rather than its
# printed 2.37, because linear disposition ties the two together: halving the
# interval doubles the steady-state average. 2 * 1.09 = 2.18, which the
# simulation reproduces to well under a percent. The printed 2.37 is
# inconsistent with the paper's other reported value, not with this model --
# see the note below. No parameter was adjusted.
stopifnot(abs(sim_cav("1.20 mg/kg q6w") - 2 * 1.09) / (2 * 1.09) < 0.02)

# ALC sits systematically a little high (all four arms high, 4-10%) because the
# packaged baseline is the Table 1 mean of 1314 while the paper's simulated
# cohort baseline was ~1351-1363. A one-sided bias of this size is the expected
# consequence of that documented choice, not a structural error.
stopifnot(median(alc_err) < 10, max(alc_err) < 15)
```

The only regimen that misses by more than rounding is 1.20 mg/kg every 6 weeks,
where the paper reports 2.37 ng/mL.
That value cannot be a steady-state average from this model, because linear
disposition forces `Cavg(q6w) = 2 * Cavg(q12w)` exactly at steady state, and the
paper's own q12w value is 1.09 ng/mL, implying 2.18 -- which is what the
simulation returns. The published 2.37 is internally inconsistent with the
paper's other reported number by the same 8%. No parameter was adjusted to close
the gap.

## Sex has little effect on ALC despite a 2.6-fold clearance difference

The paper reports "predicted ALC levels were comparable between males and
females (2723 and 2709 cells/uL, respectively, at 1.2 mg/kg every 12 weeks),
suggesting limited clinical relevance". Sex acts only on the *early* clearance
arm, so the sex-dependent elimination *rate* switches off at 350 h -- but the
difference it has already created does not vanish at that instant. With
`Vp = 15000 L` and `Q = 10.8 L/h` the peripheral compartment has a terminal
half-life of about
`r signif(log(2) / (exp(mp[["lq"]]) / exp(mp[["lvp"]])), 3)` h, so drug banked
during the first 350 h drains back slowly and the two arms converge over
several cycles rather than immediately. What the model reproduces is therefore
the paper's *conclusion* -- that a 2.6-fold early-clearance difference has
little consequence for ALC -- and the mechanism is asserted below by showing
the gap decaying cycle over cycle.

```{r sex-effect}
sex_events <- bind_rows(
  make_arm(1.2, TAU[["q12w"]], 3L, "male",   id = 1L, SEXF = 0),
  make_arm(1.2, TAU[["q12w"]], 3L, "female", id = 2L, SEXF = 1)
)

sex_raw <- rxode2::rxSolve(rxode2::zeroRe(mod), sex_events, keep = c("regimen", "tau"),
                           returnType = "data.frame", useLinCmt = FALSE)

sex_by_cycle <- bind_rows(lapply(1:3, function(cyc) {
  sex_raw |>
    filter(time >= (cyc - 1) * TAU[["q12w"]], time <= cyc * TAU[["q12w"]]) |>
    group_by(regimen) |>
    summarise(alc_avg = trapz(time, ALC) / TAU[["q12w"]], .groups = "drop") |>
    mutate(cycle = cyc)
})) |>
  pivot_wider(names_from = regimen, values_from = alc_avg) |>
  mutate(relative_gap = abs(female - male) / ((female + male) / 2))

sex_by_cycle |>
  select(cycle, male, female, relative_gap) |>
  rename("Cycle" = cycle, "Male ALC (cells/uL)" = male,
         "Female ALC (cells/uL)" = female, "Relative gap" = relative_gap) |>
  knitr::kable(digits = c(0, 0, 0, 4),
               caption = "Average ALC by sex and cycle at 1.2 mg/kg every 12 weeks; the gap decays as the peripheral reservoir drains.")

sex_gap <- sex_by_cycle$relative_gap[sex_by_cycle$cycle == 3L]
# The contrast that DRIVES the gap: early clearance is 12.40 L/h in males and
# 4.82 L/h in females.
cl_contrast <- abs(12.40 - 4.82) / mean(c(12.40, 4.82))
c(cycle3_relative_gap = sex_gap, early_cl_relative_contrast = cl_contrast,
  gap_per_unit_contrast = sex_gap / cl_contrast,
  published_gap = abs(2723 - 2709) / mean(c(2723, 2709)))

stopifnot(
  # The mechanism: the gap shrinks monotonically as the ~960 h peripheral
  # reservoir drains. This is the claim with real margin (4.9% -> 4.3% -> 2.0%).
  all(diff(sex_by_cycle$relative_gap) < 0),
  # Females have the LOWER early clearance, so the higher exposure and the
  # higher ALC throughout.
  all(sex_by_cycle$female > sex_by_cycle$male),
  # And the paper's actual conclusion: an 88% relative difference in early
  # clearance buys under 3% difference in ALC, i.e. the response is far less
  # sensitive to sex than the PK is. Achieved ratio ~0.023.
  sex_gap < 0.03,
  sex_gap / cl_contrast < 0.05
)
```

## Every evaluated regimen exceeds 1500 cells/uL

```{r above-1500}
above <- ss_summary |> filter(dose_mgkg >= 0.05)
stopifnot(nrow(above) == nrow(ARMS), all(above$alc_avg > 1500))
range(above$alc_avg)
```

# Assumptions and deviations

## Errata: three defects in the paper's printed equations

1. **The feedback ratio is inverted as typeset.** Page 9310 prints
   `(Circ / Circ0)^gamma`; the model implements `(Circ0 / Circ)^gamma`. As
   printed the drug would *reduce* ALC to a third of baseline, contradicting
   every result the paper reports. Verified numerically in "Structural checks"
   above.
2. **The last ODE prints `Ktr` where `Kcirc` belongs.** Page 9310 prints
   `dCirc/dt = Ktr * Transit3 - Ktr * Circ`, which would leave the estimated
   `Kcirc` (0.051 1/h, bootstrap 95% CI 0.026-0.158) unused anywhere in the
   model. Both the Table 3 footnote and the Figure 2 legend define `Kcirc` as
   the "degradation rate constant of circulating lymphocytes", and it is
   numerically distinct from `Ktr` = 0.0274 1/h. The prose beneath the equations
   compounds the error by calling `Kcirc` "the transit rate constant",
   contradicting the table footnote and the figure legend.
3. **`KProl` is unreported but structurally forced.** Table 3 has no `KProl`
   row. Requiring `dProl/dt = 0` at baseline gives `KProl = Ktr` identically.

## Values not reported by the paper

- **Baseline ALC (`Circ0`) is not an estimated parameter.** The PD model was
  fitted sequentially and used each subject's own observed pre-dose ALC. The
  packaged typical value is the Table 1 cohort mean of 1314.4 cells/uL. The
  Results text also gives a median of 1268 cells/uL; the mean is preferred
  because inverting the paper's own simulation ("2439-2898 cells/uL at
  0.6 mg/kg ... increases of +1088 and +1535 from baseline") implies a simulated
  cohort baseline of 1351-1363, nearer the mean. **No IIV is attached** -- the
  40.3% in Table 1 is the observed CV of the data, not an estimated omega.
  Simulated ALC in the comparison table therefore sits a few percent high.
- **The additive PK residual `epsilon_1` is reported nowhere.** The observation
  equation on page 9309 carries one, but Table 2 lists only the proportional
  term. Only the proportional term is encoded; no value was invented.
- **The endogenous IL-7 baseline typical value** is the Results-text median of
  0.05 ng/mL. Its IIV is recoverable: the paper states `var(omega_2) =
  var(epsilon_2)`, and Table 2 gives the proportional epsilon as a 27.5% CV.
  The paper's form is additive-on-one (`C_BSL * (1 + omega_2)`); the model uses
  a log-normal with the same 27.5% CV so the baseline cannot go negative.

## Simulation choices

- **Typical-value trajectories, not a stochastic cohort.** This is forced by the
  model. The saturating ceiling on ALC is `(1 + Emax)^(1/gamma)` and
  `1/gamma = 7.69`, so the published `IIV_Emax` of 3.900 (SD 1.975 on the log
  scale) is raised to the 7.69th power: a +2 SD subject has a ceiling about
  2e7 times baseline, and a 100-subject draw reaches roughly 1e8 cells/uL at the
  95th percentile. The paper itself reports this term as barely identified
  (bootstrap 95% CI 0.005-7.774, spanning three orders of magnitude), and its
  reported Monte Carlo means are reproduced by the typical-value trajectory. The
  variance is transcribed exactly as published in the model file, with a warning
  comment; it is not silently shrunk here.
- **The cohort-mean concentration correction is analytic, not fitted.** Park 2025
  reports means across a simulated cohort. `Cavg` is proportional to `1/CL`, so
  a log-normal `CL` inflates the cohort mean over the typical value by exactly
  `exp(omega^2 / 2) = exp(0.613 / 2) = 1.3587`. That factor uses only the
  published `IIV+IOV_CL` and is applied uniformly to every arm.
- **Body weight is fixed at the pooled cohort mean** of
  `r signif(WT, 3)` kg (Table 1: 67.1 kg in 19 males, 55.3 kg in 16 females).
  The model carries no allometric scaling -- clearance and volumes are absolute
  -- so mg/kg doses must be converted to an absolute mg amount, as done here.
- **`TCLchange` is measured from the start of the record**, matching the paper's
  "350 hours (14 days) after treatment initiation", not from the most recent
  dose. It has therefore fully switched over before the steady-state intervals
  analysed above.
- **`useLinCmt = FALSE`** is passed to every `rxSolve()` call for
  explicitness, not because the default is wrong here. `rxSolve()`'s default
  `useLinCmt = TRUE` is known to silently drop the peripheral compartment of a
  two-compartment model written with `k12` / `k21` micro-constants, and this
  model does write those names inside `model()`. It is **not** affected: the
  structural checks above assert that the default path and the explicit-ODE path
  agree to better than 1e-10 on both endpoints and that `peripheral1` survives,
  so a downstream user calling plain `rxSolve(readModelDb(...), ev)` gets the
  same answer this vignette reports. `ini()` stores `lq` / `lvp` rather than
  `lk12` / `lk21` for the same reason.
