---
title: "Anti-CD19 CAR-T cell therapy in systemic lupus erythematosus QSP (Park 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Anti-CD19 CAR-T cell therapy in systemic lupus erythematosus QSP (Park 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r model}
ui <- rxode2::rxode2(readModelDb("Park_2025_cd19_cart_sle_qsp"))
mod_typical <- rxode2::zeroRe(readModelDb("Park_2025_cd19_cart_sle_qsp"))
```

# Model and source

* Citation: `r ui$reference`
* Article (open access): <https://doi.org/10.1111/cts.70146>
* Appendix S1 (narrative, equations 1-53, the SLE-DAI regression output):
  `CTS-18-e70146-s003.docx`
* Appendix S2 (the complete Monolix MLXTRAN control stream):
  `CTS-18-e70146-s004.docx`
* Figures S1 (individual observed data) and S2 (goodness of fit):
  `CTS-18-e70146-s001.pdf`, `CTS-18-e70146-s002.pdf`

Both appendices are distributed through the PubMed Central open-access package
for PMC11815715.

This is a deterministic-structure quantitative systems pharmacology model with
population inter-individual variability, not a population PK model. There is no
drug concentration anywhere in it: the "dose" is a number of CAR-positive T
cells per kilogram, and the lymphodepletion chemotherapy enters through a
kinetic-pharmacodynamic (K-PD) compartment because no fludarabine or
cyclophosphamide concentrations were measured.

The authoritative source for every equation is the **Appendix S2 control
stream** - the code that actually produced the published figures. Appendix S1
prints the same system as typeset equations, and where the two disagree the
control stream is used here; each such case is listed under
[Assumptions and deviations](#assumptions-and-deviations).

## Solver settings

```{r solver}
# Converged, robust settings for this model - see "Solver behaviour" below.
AT <- 1e-6
RT <- 1e-8

solve_typical <- function(times, WT = 70, atol = AT, rtol = RT, ...) {
  ev <- rxode2::et(times)
  ev$WT <- WT
  as.data.frame(rxode2::rxSolve(mod_typical, ev, omega = NA,
                                atol = atol, rtol = rtol, ...))
}
```

The state vector spans about forty orders of magnitude: bone-marrow CAR-T pools
reach `1e11` cells/L while the eradicated CD19+ B cell pools fall below
`1e-25` cells/L. With `rxode2`'s default `atol = 1e-8` the solver demands a
step size below double-precision resolution when total CAR-T cells cross the
`CARTrev.min` reconstitution threshold late in a two-year simulation, and a
fraction of simulated subjects fail to integrate. `atol = 1e-6` with
`rtol = 1e-8` is both **converged** and **robust**. Both claims are checked
mechanically rather than asserted: [Solver behaviour](#solver-behaviour)
tightens the tolerances and asserts the reported outputs are unchanged, and the
200-subject cohort below asserts that every subject integrates and that no
output is `NA`. Use these settings for any long-horizon simulation of this
model.

# Population

```{r population}
pop <- ui$population
tibble::tibble(
  Field = names(pop),
  Value = vapply(pop, function(x) paste(as.character(x), collapse = "; "), character(1))
) |>
  knitr::kable()
```

Individual body weights are reported nowhere in Park 2025 or in either
appendix, and the clinical study that supplied the cellular-kinetic data
(Mackensen 2022, *Nat Med* 28:2124-2132) is not open access, so none could be
verified from the sources on disk. `WT` is a user-supplied covariate and every
allometric term in the model is normalized to 70 kg, which is the reference
used throughout this vignette. Supply a different weight through the `WT`
column to scale `Vb`, `Vbm`, the administered cell number, the C3 synthesis
rate and the urine volume together.

# Source trace

## Equations

Every differential equation in the model file is a transcription of the
Appendix S2 control stream; Appendix S1 numbers the same equations.

```{r eqtrace, echo = FALSE}
tibble::tribble(
  ~Component,                                   ~`Model file`,                      ~`Source location`,
  "CAR-T immunophenotypes, peripheral blood",   "d/dt(cd[48]_{tn,cm,em,ef}_pb)",    "Appendix S1 eq 1-4",
  "CAR-T immunophenotypes, bone marrow",        "d/dt(cd[48]_{tn,cm,em,ef}_bm)",    "Appendix S1 eq 5-8",
  "Free CAR / free CD19 antigen, blood",        "car_free_*_pb, ag_free_pb",        "Appendix S1 eq 9-10",
  "CAR-target complex, blood",                  "d/dt(cplx_cd[48]_pb)",             "Appendix S1 eq 11",
  "Complexes per CAR-T cell / per B cell",      "cplx_per_cart_*, cplx_per_b_*",    "Appendix S1 eq 12-13",
  "Expansion and killing Emax functions",       "kexp_*, kkill_*",                  "Appendix S1 eq 14-15",
  "Reactive CD19+ B cells, blood",              "d/dt(cd19b_pb)",                   "Appendix S1 eq 16",
  "Naive CD19+ B cell reconstitution",          "d/dt(cd19b_rev_pb), k_rev switch", "Appendix S1 eq 17-18",
  "Target engagement, bone marrow",             "ag_free_bm, d/dt(cplx_cd[48]_bm)", "Appendix S1 eq 19-25",
  "CD19+ B cells, bone marrow",                 "d/dt(cd19b_bm)",                   "Appendix S1 eq 26",
  "Lymphodepletion K-PD and Emax effects",      "d/dt(ldc), eff_cd4/cd8/nt",        "Appendix S1 eq 27-28",
  "Host T cell baselines",                      "cd4t_base, cd8t_base",             "Appendix S1 eq 29-30",
  "Host CD4+/CD8+ precursor-dependent IDR",     "d/dt(cd4t_pre), d/dt(cd4t), ...",  "Appendix S1 eq 31-34",
  "Friberg neutrophil chain",                   "ktr_nt, d/dt(nt_prol), ...",       "Appendix S1 eq 35-41",
  "Host-cell and IFN-alpha signal terms",       "stimulation / activation ratios",  "Appendix S1 eq 42-44",
  "Anti-dsDNA autoantibody release switch",     "kgen_aab, kdeg_aab, d/dt(aab)",    "Appendix S1 eq 45-47",
  "Interferon-alpha turnover",                  "ifn_rel, d/dt(ifna)",              "Appendix S1 eq 48-49",
  "Complement C3 turnover",                     "kdeg_c3, d/dt(c3)",                "Appendix S1 eq 50-51",
  "Proteinuria turnover",                       "kdeg_ptu, d/dt(ptu)",              "Appendix S1 eq 52-53",
  "SLE-DAI translation",                        "sledaiC3, sledaiIfna, sledaiAab",  "Park 2025 Figure 4A; Appendix S1 regression output"
) |>
  knitr::kable()
```

## Parameters

Every `ini()` entry carries an in-file comment naming its source location. The
table below is generated from the model object.

```{r partrace, echo = FALSE}
ui$iniDf |>
  dplyr::filter(is.na(neta1)) |>
  dplyr::transmute(
    Parameter = name,
    Estimate  = signif(est, 6),
    Fixed     = ifelse(fix, "yes", "no"),
    Label     = label
  ) |>
  knitr::kable()
```

```{r etatrace, echo = FALSE}
ui$iniDf |>
  dplyr::filter(!is.na(neta1)) |>
  dplyr::transmute(
    `Random effect` = name,
    `Variance (omega^2)` = signif(est, 6),
    `omega (Park 2025 Table 1)` = signif(sqrt(est), 4),
    Fixed = ifelse(fix, "yes", "no")
  ) |>
  knitr::kable()
```

Park 2025 Table 1 reports the random effect as omega. Appendix S1 states that
model fitting used the SAEM algorithm of Monolix 2023R1 with a log-normal IIV
distribution, so the tabulated omega is the standard deviation of the
normally-distributed eta on the log-transformed parameter; the model file
therefore stores omega^2.

## Units and dimensional analysis

```{r units, echo = FALSE}
tibble::tribble(
  ~Quantity, ~Units, ~Note,
  "Time", "day", "All rate constants are per day.",
  "CAR-T immunophenotype states", "cells/L", "Reported outputs tnCart/cmCart/emCart/efCart are the CD4+ plus CD8+ sums, matching Park 2025 Figure 2 (cells/L).",
  "totalCart, bCell, cd4t, cd8t, nt", "cells/uL", "totalCart and bCell apply the 1e-6 conversion in Appendix S2; the host-cell states are natively per uL (Circ0.NT = 5617 cells/uL).",
  "CAR-target complexes", "number/L", "kon converted from 1/(M s) by 86400 / 6.023e23; koff from 1/s by 86400.",
  "Complexes per cell", "number/cell", "Drives EC50exp (1.9e-5 per effector cell) and KC50 (7.71 per B cell).",
  "ldc", "mg", "K-PD amount; c_ldc = ldc / Vb in mg/L, matched to EC50 values in mg/L.",
  "aab", "IU/L", "Ksyn.AAb is IU/cell/day; multiplied by cells/L gives IU/L/day.",
  "ifna", "pg/L", "State is pg/L; the reported output ifnaConc = ifna / 1000 is pg/mL, as plotted in Figure S1E.",
  "c3", "mg/L", "Synthesis 1.5 mg/kg/h x 24 h x WT kg = mg/day, divided by Vb (L) gives mg/L/day.",
  "ptu", "mg/g creatinine", "Excretion 1000 mg/day divided by Vu (dL/day); the paper reports urine protein per gram creatinine and treats the baseline regressor as the unit anchor.",
  "Vb, Vbm", "L", "5 and 1.6 L at 70 kg, scaled linearly with WT.",
  "Vu", "dL/day", "1 cc/kg/h = WT x 24 / 100 dL/day."
) |>
  knitr::kable()
```

The body-weight dependence cancels in the complement C3 turnover rate:
`ksyn_c3_day / (rbase_c3 * vb) = (1.5 * 24 * WT) / (rbase_c3 * 5 * WT / 70)`
`= 504 / rbase_c3` per day, independent of WT. It does **not** cancel in the
proteinuria turnover rate, `ksyn_ptu / (vu * rbase_ptu)`, which is inversely
proportional to body weight.

# Validation

`PKNCA`-based non-compartmental analysis is not applicable: there is no
concentration-time profile of an administered drug. The checks below follow the
mechanistic-model pattern - baseline flux balance, steady-state hold,
perturbation recovery, closed-form identities, and replication of the published
figures.

## 1. Baseline flux balance

Four of the model's turnover states are constructed so that the pre-treatment
baseline is an exact steady state: the degradation rate constant is
back-calculated from the synthesis rate and the baseline. This is checked
symbolically, from the `ini()` values, without solving the ODEs.

```{r flux}
p <- setNames(ui$iniDf$est, ui$iniDf$name)
lin <- function(nm) if (startsWith(nm, "l")) exp(p[[nm]]) else p[[nm]]

WT      <- 70
vb      <- lin("vb_ref") * WT / 70
vu      <- WT * 24 / 100
b0      <- lin("lrbase_cd19b")
aab0    <- lin("lrbase_aab")
c30     <- lin("lrbase_c3")
ifna0   <- lin("lrbase_ifna") * 1000
ptu0    <- lin("lrbase_ptu")
nt0     <- lin("lrbase_nt")

flux <- tibble::tribble(
  ~State, ~`d/dt at baseline`, ~`Steady state?`,
  "cd4t",
  (lin("lk0_cd4pre") / p[["kprol_cd4pre"]]) * p[["kprol_cd4pre"]] -
    (lin("lk0_cd4pre") / lin("lkout_cd4")) * lin("lkout_cd4"),
  "yes, by construction (Appendix S1 eq 29, 31-32)",
  "cd8t",
  (lin("lk0_cd8pre") / p[["kprol_cd8pre"]]) * p[["kprol_cd8pre"]] -
    (lin("lk0_cd8pre") / lin("lkout_cd8")) * lin("lkout_cd8"),
  "yes, by construction (Appendix S1 eq 30, 33-34)",
  "ifna",
  p[["n_pdc"]] * (ifna0 * lin("lkdeg_ifna") / p[["n_pdc"]]) - ifna0 * lin("lkdeg_ifna"),
  "yes, by construction (Appendix S1 eq 48-49)",
  "aab",
  b0 * lin("lksyn_aab") - aab0 * (b0 * lin("lksyn_aab") / aab0),
  "yes, by construction (Appendix S1 eq 46-47)",
  "c3",
  (p[["ksyn_c3"]] * 24 * WT -
     (p[["ksyn_c3"]] * 24 * WT / (c30 * vb)) * c30 * vb) / vb,
  "yes, by construction (Appendix S1 eq 50-51)",
  "cd19b_pb",
  p[["kg_cd19b"]] * b0,
  "no - the reactive pool has no homeostatic loss term and grows at Kg in the absence of CAR-T killing",
  "ptu",
  (p[["ksyn_ptu"]] * exp(1) * exp(1) - ptu0 * vu * (p[["ksyn_ptu"]] / (vu * ptu0))) / vu,
  "no - Appendix S2 applies the CD8+ T cell and neutrophil signals as exp(x/x0), which equals e rather than 1 at baseline"
)
flux |>
  dplyr::mutate(`d/dt at baseline` = signif(`d/dt at baseline`, 4)) |>
  knitr::kable()

stopifnot(
  abs(flux$`d/dt at baseline`[1:5]) <
    c(1e-9, 1e-9, 1e-9, 1e-3, 1e-9) * c(1, 1, ifna0, aab0, c30)
)
```

The two states that are *not* at steady state are properties of the published
model, not of this transcription; both are discussed under
[Assumptions and deviations](#assumptions-and-deviations).

## 2. Steady-state hold of the host immune cells

Setting the lymphodepletion amount to zero removes the only forcing on the host
CD4+ / CD8+ T cell and neutrophil subsystems. They must then hold at their
constructed baselines for the full two years.

```{r hold}
hold <- solve_typical(seq(0, 730, by = 1), params = c(dose_ldc = 0))

drift <- c(
  cd4t = max(abs(hold$cd4t / hold$cd4t[1] - 1)),
  cd8t = max(abs(hold$cd8t / hold$cd8t[1] - 1)),
  nt   = max(abs(hold$nt   / hold$nt[1]   - 1))
)
tibble::tibble(
  State = names(drift),
  `Baseline` = signif(c(hold$cd4t[1], hold$cd8t[1], hold$nt[1]), 6),
  `Expected (K0/Kout, Circ0)` = signif(
    c(lin("lk0_cd4pre") / lin("lkout_cd4"),
      lin("lk0_cd8pre") / lin("lkout_cd8"),
      nt0), 6),
  `Max relative drift over 730 days` = signif(drift, 3)
) |>
  knitr::kable()

stopifnot(drift < 1e-9)
stopifnot(abs(hold$cd4t[1] / (lin("lk0_cd4pre") / lin("lkout_cd4")) - 1) < 1e-10)
stopifnot(abs(hold$cd8t[1] / (lin("lk0_cd8pre") / lin("lkout_cd8")) - 1) < 1e-10)
stopifnot(abs(hold$nt[1] / nt0 - 1) < 1e-10)
```

## 3. Perturbation recovery of the neutrophil chain

Displacing the circulating neutrophil pool to half its baseline (an `evid = 5`
replacement record on the `nt` compartment, since the model sets the initial
condition in `model()` and would otherwise ignore `inits =`) must trigger the
Friberg feedback term `(Circ0/NT)^gamma` and drive the pool back to baseline.

A transit chain with proliferative feedback is a *damped oscillator*, not a
monotone relaxation: the feedback term over-corrects while the depleted signal
is still propagating down the three transit compartments, so the pool
overshoots baseline once and then rings back down. The checks below therefore
assert the properties the Friberg structure actually implies - a single
overshoot of modest amplitude, a decaying oscillation, and exact return to
baseline - rather than monotonicity. The same overshoot is visible in the
observed data: the neutrophil panel of Figure S1C rebounds above each patient's
pre-treatment level around day 15-30 before settling.

```{r perturb, fig.height = 3.2}
ev_pert <- rxode2::et(amt = 0.5 * nt0, cmt = "nt", evid = 5, time = 0) |>
  rxode2::et(seq(0, 90, by = 0.5))
ev_pert$WT <- 70
pert <- as.data.frame(rxode2::rxSolve(mod_typical, ev_pert, omega = NA,
                                      atol = AT, rtol = RT,
                                      params = c(dose_ldc = 0)))

ggplot(pert, aes(time, nt)) +
  geom_hline(yintercept = nt0, linetype = "dashed") +
  geom_line(linewidth = 0.8) +
  labs(x = "Time (day)", y = "Neutrophils (cells/uL)",
       title = "Recovery from a 50% neutrophil depletion, no lymphodepletion",
       subtitle = "Dashed line: constructed baseline Circ0") +
  theme_bw()

recovery_err <- abs(pert$nt[nrow(pert)] / nt0 - 1)
overshoot <- max(pert$nt) / nt0
# Amplitude of the first ring above baseline vs the second, as a decay check.
above <- pert$nt - nt0
ring1 <- max(above[pert$time > 0 & pert$time <= 20])
ring2 <- max(above[pert$time > 30])
cat(sprintf(
  "nt(0) = %.1f, peak = %.1f (%.3fx baseline) at day %.1f, nt(90 d) = %.1f,\nbaseline = %.1f, relative error = %.2g, ring decay = %.3g -> %.3g\n",
  pert$nt[1], max(pert$nt), overshoot, pert$time[which.max(pert$nt)],
  pert$nt[nrow(pert)], nt0, recovery_err, ring1, ring2))

stopifnot(pert$nt[1] < 0.51 * nt0)         # perturbation applied
stopifnot(recovery_err < 1e-4)             # returns to baseline
stopifnot(overshoot > 1, overshoot < 1.2)  # single modest overshoot
stopifnot(ring2 < 0.1 * ring1)             # oscillation is damped
stopifnot(min(pert$nt) == pert$nt[1])      # never falls below the perturbed start
```

## 4. Closed-form identity for the complement C3 plateau

After the reactive B cell pool is eradicated, anti-dsDNA release switches off
(Appendix S1 eq 45) and `aab` freezes. The C3 equation
`d/dt(c3) = (Ksyn - Kdeg * c3 * Vb * (aab/aab0)^gammaC3) / Vb` then has the
exact fixed point

```
c3_plateau = c3_0 * (aab_0 / aab_plateau)^gammaC3
```

which is a strong, parameter-free check on the coupling between the
autoantibody and complement layers.

```{r c3plateau}
base <- solve_typical(seq(0, 730, by = 1))

aab_plateau <- base$aab[nrow(base)]
c3_closed   <- c30 * (aab0 / aab_plateau)^lin("lgamma_c3")

tibble::tibble(
  Quantity = c("aab plateau (IU/L)", "C3 plateau, simulated (mg/L)",
               "C3 plateau, closed form (mg/L)", "Relative error"),
  Value = c(signif(aab_plateau, 6), signif(base$c3[nrow(base)], 6),
            signif(c3_closed, 6),
            signif(abs(base$c3[nrow(base)] / c3_closed - 1), 3))
) |>
  knitr::kable()

stopifnot(abs(base$c3[nrow(base)] / c3_closed - 1) < 1e-10)
```

The simulated plateau, 1111 mg/L from a baseline of 555 mg/L, sits inside the
800-2000 mg/L band of individual model fits in Park 2025 Figure 2 and matches
the roughly 1100 mg/L median prediction of Figure 3B2.

## 5. Closed-form identity for the proteinuria decline

Once the CD8+ T cell, neutrophil and interferon-alpha signals collapse after
lymphodepletion and B cell eradication, the proteinuria equation reduces to a
first-order decay whose rate constant depends only on the excretion rate, the
urine volume and the patient's own baseline:

```
PTU(t) / PTU_0 = exp(-Ksyn.PTU * t / (Vu * PTU_0))
```

This identity is what discriminates the control-stream value
`Ksyn.PTU = 1000 mg/day` from the `100 mg/day` printed in Park 2025 Table 1: at
100 mg/day a patient starting from 8100 mg/g creatinine would still be at 94%
of baseline on day 90, whereas Figure 2 shows that patient at roughly half of
baseline.

```{r ptu}
ptu_baselines <- c(8100, 6570, 3060, 2010, 75)  # digitized, Park 2025 Figure S1E

ptu_tab <- lapply(ptu_baselines, function(b0v) {
  s <- solve_typical(seq(0, 90, by = 1), params = c(lrbase_ptu = log(b0v)))
  tibble::tibble(
    `PTU baseline (mg/g creatinine)` = b0v,
    `Simulated day 90` = signif(s$ptu[s$time == 90], 4),
    `Closed form day 90` = signif(b0v * exp(-p[["ksyn_ptu"]] * 90 / (vu * b0v)), 4),
    `Ksyn.PTU = 100 mg/day` = signif(b0v * exp(-100 * 90 / (vu * b0v)), 4)
  )
}) |>
  dplyr::bind_rows()

knitr::kable(ptu_tab)

# The highest-baseline patient is the one that can be read reliably off the
# Park 2025 Figure 2 proteinuria panel: 8100 -> about 4200 mg/g creatinine.
sim_8100 <- ptu_tab$`Simulated day 90`[ptu_tab$`PTU baseline (mg/g creatinine)` == 8100]
stopifnot(abs(sim_8100 / 4200 - 1) < 0.05)
stopifnot(all(ptu_tab$`Simulated day 90` <= ptu_tab$`PTU baseline (mg/g creatinine)`))
```

The simulated day-90 value for the 8100 mg/g creatinine patient,
`r signif(sim_8100, 4)`, is within 5% of the roughly 4200 read off Figure 2,
while the Table 1 value of 100 mg/day would predict about 7600. The simulated
values run slightly above the pure closed form because the `exp(CD8T/CD8T_0)`
and `exp(NT/NT_0)` factors are `e` rather than 1 in the first day or two, before
lymphodepletion takes hold.

# Replication of the published figures

## Park 2025 Figure 2 - individual fits over 90 days

```{r fig2, fig.height = 8}
d90 <- solve_typical(seq(0, 90, by = 0.25))

panels <- d90 |>
  dplyr::transmute(
    time,
    `Total CAR-T (cells/uL)`          = totalCart,
    `CD19+ B cells (cells/uL)`        = bCell,
    `Naive CAR-T (cells/L)`           = tnCart,
    `Central memory CAR-T (cells/L)`  = cmCart,
    `Effector memory CAR-T (cells/L)` = emCart,
    `Effector CAR-T (cells/L)`        = efCart,
    `CD4+ T cells (cells/uL)`         = cd4t,
    `CD8+ T cells (cells/uL)`         = cd8t,
    `Neutrophils (cells/uL)`          = nt,
    `Interferon-alpha (pg/mL)`        = ifnaConc,
    `Anti-dsDNA (IU/L)`               = aab,
    `Complement C3 (mg/L)`            = c3,
    `Proteinuria (mg/g creatinine)`   = ptu
  ) |>
  tidyr::pivot_longer(-time, names_to = "output", values_to = "value") |>
  dplyr::mutate(output = factor(output, levels = unique(output)))

ggplot(panels, aes(time, value)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~output, scales = "free_y", ncol = 3) +
  scale_y_log10() +
  labs(x = "Time (day)", y = NULL,
       title = "Typical-value profiles, 1e6 CAR-T cells/kg in a 70 kg patient",
       subtitle = "Replicates Park 2025 Figure 2 (which shows the five individual fits)") +
  theme_bw() +
  theme(strip.text = element_text(size = 7))
```

The typical-value trajectory reproduces every qualitative feature the paper
describes, and each landmark falls inside the spread of the five individual fits
in Figure 2.

```{r fig2check, echo = FALSE}
tibble::tribble(
  ~Feature, ~`Model (typical)`, ~`Park 2025 Figure 2 / S1 (5 patients)`,
  "Total CAR-T peak (cells/uL)",
  signif(max(d90$totalCart), 4),
  "about 30 to 1000",
  "Time of total CAR-T peak (day)",
  d90$time[which.max(d90$totalCart)],
  "7 to 10",
  "Naive CAR-T peak (cells/L)",
  signif(max(d90$tnCart), 3),
  "about 5e5 to 1e7",
  "CD4+ T cell nadir (cells/uL)",
  signif(min(d90$cd4t), 3),
  "about 20 to 100, around day 3",
  "CD8+ T cell nadir (cells/uL)",
  signif(min(d90$cd8t), 3),
  "about 10 to 30, around day 3",
  "Neutrophil nadir (cells/uL)",
  signif(min(d90$nt), 3),
  "about 200 to 2000, day 7 to 9",
  "Interferon-alpha day-90 (pg/mL)",
  signif(d90$ifnaConc[nrow(d90)], 3),
  "about 0.02 to 0.3",
  "Complement C3 plateau (mg/L)",
  signif(d90$c3[nrow(d90)], 4),
  "about 800 to 2000"
) |>
  knitr::kable()

stopifnot(
  max(d90$totalCart) > 30, max(d90$totalCart) < 1000,
  d90$time[which.max(d90$totalCart)] >= 7, d90$time[which.max(d90$totalCart)] <= 10,
  min(d90$cd4t) > 20, min(d90$cd4t) < 100,
  min(d90$cd8t) > 10, min(d90$cd8t) < 30,
  d90$c3[nrow(d90)] > 800, d90$c3[nrow(d90)] < 2000
)
```

## Park 2025 Figures 3 and 5 - two-year population prediction

The paper simulates two years post-infusion with 95% confidence intervals and
compares against the long-term follow-up cohort. Here 200 subjects are drawn
from the published IIV.

```{r fig3, fig.height = 6}
set.seed(20250213)
ev_pop <- rxode2::et(seq(0, 730, by = 2))
ev_pop$WT <- 70
sim_pop <- rxode2::rxSolve(readModelDb("Park_2025_cd19_cart_sle_qsp"), ev_pop,
                           nSub = 200, atol = AT, rtol = RT) |>
  as.data.frame()

# rxSolve names the subject column sim.id when subjects are generated internally
id_col <- if ("id" %in% names(sim_pop)) "id" else "sim.id"
stopifnot(length(unique(sim_pop[[id_col]])) == 200L)
stopifnot(!anyNA(sim_pop$totalCart), !anyNA(sim_pop$bCell), !anyNA(sim_pop$c3))

bands <- sim_pop |>
  dplyr::transmute(
    time,
    `Total CAR-T (cells/uL)`        = totalCart,
    `CD19+ B cells (cells/uL)`      = bCell,
    `CD4+ T cells (cells/uL)`       = cd4t,
    `Neutrophils (cells/uL)`        = nt,
    `Anti-dsDNA (IU/L)`             = aab,
    `Complement C3 (mg/L)`          = c3,
    `Proteinuria (mg/g creatinine)` = ptu
  ) |>
  tidyr::pivot_longer(-time, names_to = "output", values_to = "value") |>
  dplyr::group_by(output, time) |>
  dplyr::summarise(
    med = median(value),
    lo  = quantile(value, 0.025),
    hi  = quantile(value, 0.975),
    .groups = "drop"
  ) |>
  dplyr::mutate(output = factor(output, levels = unique(output)))

ggplot(bands, aes(time, med)) +
  geom_ribbon(aes(ymin = pmax(lo, 1e-3), ymax = hi), alpha = 0.25) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~output, scales = "free_y", ncol = 3) +
  scale_y_log10() +
  labs(x = "Time (day)", y = NULL,
       title = "Two-year population prediction, median and 95% interval (n = 200)",
       subtitle = "Replicates Park 2025 Figure 3A/3B and Figure 5A") +
  theme_bw() +
  theme(strip.text = element_text(size = 8))
```

```{r fig3check, echo = FALSE}
peak <- tapply(sim_pop$totalCart, sim_pop[[id_col]], max)
b730 <- sim_pop$bCell[sim_pop$time == 730]
tibble::tibble(
  Quantity = c("Peak total CAR-T (cells/uL)", "CD19+ B cells at day 730 (cells/uL)"),
  Median   = signif(c(median(peak), median(b730)), 4),
  `2.5th`  = signif(c(quantile(peak, 0.025), quantile(b730, 0.025)), 4),
  `97.5th` = signif(c(quantile(peak, 0.975), quantile(b730, 0.975)), 4),
  `Park 2025` = c("individual peaks about 30 to 3000 (Figure 3A2)",
                  "about 200 to 900 (Figure 3A3)")
) |>
  knitr::kable()
```

The model reproduces the paper's central narrative: the endogenous-lymphocyte
nadir coincides with the CAR-T expansion peak, maximum CD19+ B cell depletion
occurs at the CAR-T Cmax, and B cells reconstitute over the following year
exclusively through the naive (non-autoreactive) pool.

## Park 2025 Figure 6 - dose-ranging simulations

```{r fig6, fig.height = 4.5}
dose_levels <- c(0.01, 0.05, 0.10, 0.50, 1.00)

dose_sim <- lapply(dose_levels, function(f) {
  s <- solve_typical(seq(0, 200, by = 0.5), params = c(dose_cart = 1e6 * f))
  s$dose_label <- sprintf("%g%% of 1e6 cells/kg", f * 100)
  s
}) |>
  dplyr::bind_rows() |>
  dplyr::mutate(dose_label = factor(dose_label,
                                    levels = sprintf("%g%% of 1e6 cells/kg",
                                                     dose_levels * 100)))

dose_sim |>
  dplyr::select(time, dose_label,
                `Total CAR-T (cells/uL)` = totalCart,
                `CD19+ B cells (cells/uL)` = bCell) |>
  tidyr::pivot_longer(c(-time, -dose_label), names_to = "output", values_to = "value") |>
  ggplot(aes(time, value, colour = dose_label)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~output, scales = "free_y") +
  scale_y_log10() +
  scale_colour_viridis_d(end = 0.9) +
  labs(x = "Time (day)", y = NULL, colour = "CAR-T dose",
       title = "Dose-ranging simulation",
       subtitle = "Replicates Park 2025 Figure 6A") +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r fig6check, echo = FALSE}
dose_tab <- dose_sim |>
  dplyr::group_by(dose_label) |>
  dplyr::summarise(
    `Peak total CAR-T (cells/uL)` = signif(max(totalCart), 4),
    `Tmax (day)` = time[which.max(totalCart)],
    `Days to 50% B cell depletion` = time[which(bCell < 0.5 * bCell[1])[1]],
    `CD19+ B cells at day 200 (cells/uL)` = signif(bCell[which.max(time)], 4),
    .groups = "drop"
  ) |>
  dplyr::rename(`CAR-T dose` = dose_label)

knitr::kable(dose_tab)

# Park 2025 Results 3.5: "decline in dose level leads to a saturable
# under-proportional cellular expansion, delayed Cmax (or prolonged Tmax) but
# overall, very similar persistence values", and "at very low doses ... there
# seems to be a delayed effect in elimination of all the reactive CD19+ B cell
# population" without a material effect on naive B cell regeneration.
peaks <- dose_tab$`Peak total CAR-T (cells/uL)`
tmaxs <- dose_tab$`Tmax (day)`
dep   <- dose_tab$`Days to 50% B cell depletion`
b200  <- dose_tab$`CD19+ B cells at day 200 (cells/uL)`

stopifnot(peaks[length(peaks)] / peaks[1] < 100 * 0.5)     # 100-fold dose -> < 50-fold peak
stopifnot(tmaxs[1] > tmaxs[length(tmaxs)])                 # delayed Cmax at lower doses
stopifnot(dep[1] > dep[length(dep)])                       # delayed B cell depletion
stopifnot(max(b200) / min(b200) < 2)                       # similar naive B cell regeneration
```

A 100-fold dose reduction lowers the CAR-T peak only about
`r signif(peaks[length(peaks)] / peaks[1], 3)`-fold, delays Tmax from
`r tmaxs[length(tmaxs)]` to `r tmaxs[1]` days, and delays 50% B cell depletion
from `r dep[length(dep)]` to `r dep[1]` days, while day-200 naive B cell
regeneration is essentially unchanged - exactly the behaviour reported in
Park 2025 Results 3.5.

## Park 2025 Figure 4B - translation to the SLE disease activity index

```{r fig4, fig.height = 3.4}
sledai <- d90 |>
  dplyr::select(time,
                `C3-based (Figure 4B1)`         = sledaiC3,
                `IFN-alpha-based (Figure 4B2)`  = sledaiIfna,
                `Anti-dsDNA-based (Figure 4B3)` = sledaiAab) |>
  tidyr::pivot_longer(-time, names_to = "biomarker", values_to = "SLEDAI")

ggplot(sledai, aes(time, SLEDAI)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~biomarker) +
  labs(x = "Time (day)", y = "SLE-DAI",
       title = "Biomarker to SLE-DAI translation",
       subtitle = "Replicates Park 2025 Figure 4B") +
  theme_bw()

tibble::tibble(
  Biomarker = c("Complement C3", "Interferon-alpha", "Anti-dsDNA"),
  `SLE-DAI day 0`  = signif(c(d90$sledaiC3[1], d90$sledaiIfna[1], d90$sledaiAab[1]), 4),
  `SLE-DAI day 90` = signif(c(d90$sledaiC3[nrow(d90)], d90$sledaiIfna[nrow(d90)],
                              d90$sledaiAab[nrow(d90)]), 4),
  `Park 2025 Figure 4B` = c("about 10 falling to a plateau near 7",
                            "high early, plateau near 4",
                            "high early, plateau near 17 (see Errata)")
) |>
  knitr::kable()

# Figure 4B1: the C3-based translation is the one the paper endorses.
stopifnot(d90$sledaiC3[1] > 8, d90$sledaiC3[1] < 14)
stopifnot(d90$sledaiC3[nrow(d90)] > 4, d90$sledaiC3[nrow(d90)] < 8)
```

Park 2025 Results 3.3 concludes that "predictions based on the C3 protein
(Figure 4B1) provided the best approximation of the SLE-DAI", and that the
interferon-alpha and anti-dsDNA regressions have y-intercepts well above zero
that limit their usefulness. The model reproduces that ordering.

# Solver behaviour {#solver-behaviour}

```{r converge}
grid <- expand.grid(atol = c(1e-4, 1e-6, 1e-8), rtol = c(1e-6, 1e-8))
conv <- lapply(seq_len(nrow(grid)), function(i) {
  s <- solve_typical(seq(0, 730, by = 1), atol = grid$atol[i], rtol = grid$rtol[i])
  tibble::tibble(
    atol = grid$atol[i], rtol = grid$rtol[i],
    # kable would print these as 0e+00 if left numeric
    `atol ` = format(grid$atol[i], scientific = TRUE),
    `rtol ` = format(grid$rtol[i], scientific = TRUE),
    `Peak CAR-T (cells/uL)` = signif(max(s$totalCart), 7),
    `C3 day 90 (mg/L)` = signif(s$c3[s$time == 90], 7),
    `Proteinuria day 90` = signif(s$ptu[s$time == 90], 7)
  )
}) |>
  dplyr::bind_rows()

knitr::kable(conv |> dplyr::select(-atol, -rtol))

# Convergence gate: the settings used throughout this vignette must agree with
# a hundredfold-tighter absolute tolerance on every reported output.
ref  <- conv[conv$atol == 1e-8 & conv$rtol == 1e-8, -(1:4)]
used <- conv[conv$atol == AT   & conv$rtol == RT,   -(1:4)]
reldiff <- abs(as.numeric(used) / as.numeric(ref) - 1)
names(reldiff) <- names(ref)
print(signif(reldiff, 3))
stopifnot(all(reldiff < 1e-5))
```

The settings used throughout this vignette (`atol = 1e-6`, `rtol = 1e-8`) agree
with the hundredfold-tighter `atol = 1e-8, rtol = 1e-8` to better than one part
in 1e5 on every reported output - the `stopifnot` above is what enforces that, so the
claim cannot silently rot. Loosening the absolute tolerance to `1e-4` visibly
moves the anti-dsDNA plateau, which is set at a switch crossing (Errata 13).
Unlike the rxode2 defaults, these settings also integrate all 200 simulated
subjects in the cohort above without failure, which that chunk asserts
directly.

# Assumptions and deviations {#assumptions-and-deviations}

**Conflicts between the two appendices, resolved in favour of the control
stream (Appendix S2), which is the code that produced the published figures.**

1. **Urinary protein excretion rate.** Park 2025 Table 1 prints
   `K.PTU.syn = 100 mg/day`; Appendix S2 sets `PTU_Ksyn = 1000/1` with the
   inline comment `1000 mg/day`. The model file uses 1000. Validation section 5
   shows that only the control-stream value reproduces the proteinuria decline
   in Figure 2; at 100 mg/day the profile would be essentially flat over
   90 days.
2. **Proteinuria signal terms.** Appendix S1 eq 44 prints the CD8+ T cell and
   neutrophil signals as bare ratios `CD8T/CD8T_0` and `NT/NT_0`; Appendix S2
   applies them as `exp(CD8T/CD8T_0)` and `exp(NT/NT_0)`. The model file uses
   the exponential form. The two differ materially only in the first day or two
   after infusion, before lymphodepletion collapses both signals; the long-term
   decline is identical.
3. **Direction of the C3 drive on proteinuria.** Appendix S1 eq 52 prints the
   complement term as `(C3/C3_0)^gamma.PTU`; Appendix S2 codes it inverted, as
   `(C3_0/C3)^PTUgam`. Only the control-stream direction is mechanistically
   coherent and only it reproduces the published figures. C3 is *depressed* in
   active SLE and rises back toward normal as treatment removes the
   immune-complex load, so a term in `C3/C3_0` would make proteinuria climb as
   the patient improves, contradicting both Figure 2 and the stated mechanism
   ("protein excretion ... eventually ameliorated following the reduced
   metabolism of C3", Figure 1B caption). The model file uses `(C3_0/C3)`.
4. **Naive B cell reconstitution switch.** Appendix S1 eq 17 adds a
   `- K_kill * CD19B` term to the reconstitution equation and eq 18 gates the
   switch on `CD19_total`. Appendix S2 has neither: it gates on total CAR-T
   cells and has no killing term on the naive pool. Park 2025 Table 1 describes
   `CARTrev.min` as a "threshold level of total CAR-T cells", which agrees with
   the control stream, so the control stream is used.
5. **`CARTrev.min` units.** Table 1 labels this parameter cells/L, but the
   control stream compares it against `TotCART_PB`, which carries the `1e-6`
   conversion to cells/uL. It is encoded here as 2.38 cells/uL.
6. **Interferon-alpha release rate.** Table 1 reports
   `K.IFNa.rel = 0.013 pg/cell/day` as an estimated parameter, but in Appendix
   S2 this quantity is not estimated - it is derived as
   `IFNa_0 * Kdeg_IFN / N_pDC` so that the individual baseline is a steady
   state. The derived form is used. Because the entire interferon-alpha
   subsystem depends only on `IFNa/IFNa_0`, the absolute baseline sets the
   reported scale but has no effect on any other state. Note that the tabulated
   0.013 pg/cell/day corresponds to a baseline of 0.647 pg/mL, which is at the
   very bottom of the 0.3-105 pg/mL range observed in Figure S1E.

**Values that are not published as such.**

7. **Individual baseline regressors.** `dsDNA0`, `INFA0`, `C30` and `PTU0` are
   per-subject regressors in the Monolix dataset (Appendix S2 input block) and
   are not tabulated anywhere. The defaults in the model file - 470000 IU/L,
   21 pg/mL, 555 mg/L and 3060 mg/g creatinine - are the medians of the five
   modelled patients read off Park 2025 Figure S1 panel E, and every one of them
   is flagged in the model file as figure-digitized. Reading `CD19B_PB_int` the
   same way gives 145 cells/uL against the 143 cells/uL that Table 1 reports for
   the same quantity, which is the accuracy this digitization achieves. Change
   them through `rxSolve(params = ...)` to simulate a specific patient.
8. **Lymphodepletion amount.** The K-PD compartment receives a dose whose size
   is nowhere reported. It is reconstructed from the regimen stated in the Park
   2025 Discussion - fludarabine 25 mg/m^2^/day on three days plus
   cyclophosphamide 1000 mg/m^2^ - at a rounded 1.8 m^2^ reference body surface
   area, giving 1935 mg. The host-cell response is insensitive to this choice
   over any plausible range because `EC50.CD4` (0.15 mg/L) and `EC50.CD8`
   (0.067 mg/L) are more than three orders of magnitude below the initial K-PD
   concentration, and `Emax.NT` (27.32) drives the neutrophil term far past
   full inhibition either way. Change it through `rxSolve(params = c(dose_ldc = ...))`.
9. **Lymphodepletion timing.** Patients received lymphodepletion on days -5 to
   -3 relative to infusion, but Appendix S2 sets the ODE initial time to zero
   with all host cells at their pre-treatment baselines, so the control stream
   administers lymphodepletion and CAR-T cells at the same model time origin.
   This transcription keeps that simplification; it is the reason the simulated
   host-cell nadirs sit two to three days later in model time than they would
   on the clinical calendar.
10. **No residual error model.** The paper states that residual errors were
   estimated but reports no residual-error values in Table 1 or either appendix.
   Rather than invent them, the model file carries none, and the IIV that *is*
   published is encoded in full.

**Properties of the published model worth knowing before using it.**

11. **The reactive CD19+ B cell pool has no homeostatic loss.** Its only
    equation is `d/dt = Kg * B - Kkill * B`, so in the absence of CAR-T cells it
    grows exponentially at 1.93% per day. The model is meaningful only from the
    moment of CAR-T infusion onward; it does not describe an untreated patient.
12. **The bone-marrow CAR-T pools have no elimination term** (Appendix S1
    eq 5-8 and the control stream both omit one), so they accumulate to
    approximately 1e11 cells/L and drain back to blood only through
    `K21 = 0.0017/day`. The marrow pool is not an observed output of the paper.
13. **Anti-dsDNA turnover is effectively instantaneous.** `Kdeg.AAb` is derived
    as `CD19B_PB_0 * Ksyn.AAb / AAb_0`, which for the tabulated values is about
    5e6 per day, so `aab` tracks the reactive B cell pool with no lag. When the
    pool falls below 1 cell/uL, Appendix S1 eq 45 switches release off - which
    also zeroes the derived degradation rate - and `aab` freezes at whatever
    value it then holds. That frozen floor is the flat plateau visible in the
    Figure 2 anti-dsDNA panel, and it makes the plateau slightly dependent on
    solver tolerance because it is set at a switch crossing.
14. **Reconstituted naive B cells are not CAR-T targets.** `cd19b_rev_pb` never
    enters the free-antigen balance, so the reconstituting pool is invisible to
    the remaining CAR-T cells. This is what the paper intends by "nonreactive"
    naive B cells, but it means the model cannot describe a second depletion
    wave.
15. **Anti-dsDNA to SLE-DAI translation.** The Appendix S1 regression output
    gives intercept 6.96 and slope 0.05 for the literature (`SLEDAI.Ref`) fit,
    which the model file uses with anti-dsDNA in IU/mL as labelled on the
    Figure 4A3 axis. The red regression line drawn in Figure 4A3 has a visual
    intercept near 4.3 and a slope near 4.8 per axis unit, roughly a hundredfold
    steeper, so the units of the predictor in that panel cannot be reconciled
    with the printed coefficients from the sources on disk. The tabulated
    coefficients are used because they are the numeric source; the resulting
    day-0 score of about 30 is above the range plotted in Figure 4B3. The C3 and
    interferon-alpha panels have no such discrepancy - their printed
    coefficients reproduce the plotted lines exactly - and the paper itself
    identifies C3 as the reliable translator.

**Numerical safeguards added in this transcription.**

16. `max(x, 0)` guards on free-CAR, free-antigen and complex quantities, and a
    1e-12 floor on the states that appear in fractional powers or in
    denominators. Every guarded quantity is non-negative in exact arithmetic,
    so none of them changes the published solution; they exist because the
    eradicated B cell pools decay past the point where a scalar absolute
    tolerance can distinguish them from round-off. The typical-value trajectory
    is bit-identical with and without them.
17. The solver settings documented above (`atol = 1e-6`, `rtol = 1e-8`) are a
    requirement of this model, not a preference. See
    [Solver behaviour](#solver-behaviour).

# Session info

```{r session, echo = FALSE}
sessionInfo()
```
