---
title: "Sapanisertib concentration-QTc and concentration-RR (Patel 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sapanisertib concentration-QTc and concentration-RR (Patel 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r readModelDb("Patel_2020_sapanisertib_QTcI")()$reference`
* Article: <https://doi.org/10.1002/cpdd.808>

Patel et al. (2020) report a dedicated QTc (DQT) study of the mTORC1/2 inhibitor
sapanisertib (MLN0128 / INK128 / TAK-228) in 44 patients with advanced solid
tumors, each given a single supratherapeutic 40 mg oral dose with time-matched
day -1 baseline Holter electrocardiograms. Three linear mixed-effects
concentration-effect models were fitted independently, one per electrocardiographic
endpoint, and are packaged here as three model files sharing one vignette:

| Model file | Endpoint | Source panel |
|------------|----------|--------------|
| `Patel_2020_sapanisertib_QTcI` | change from time-matched baseline in the individually rate-corrected QT interval (`QTcI`, msec) | Figure 4A |
| `Patel_2020_sapanisertib_QTcF` | change from time-matched baseline in the Fridericia-corrected QT interval (`QTcF`, msec) | Figure 4C |
| `Patel_2020_sapanisertib_RR` | change from time-matched baseline in the RR interval (`dRR`, msec) | Supplementary Figure S2 |

All three share the identical algebraic structure

$$\Delta\text{endpoint} = e_0 + \text{slope} \times C_\text{sapanisertib}$$

and all three are PD-only: the source paper fits **no** population PK model, so
sapanisertib plasma concentration enters as the time-varying covariate
`CP_SAPANISERTIB_NGML` (ng/mL) rather than being computed from an internal
compartment system.

The paper's central finding is negative, and the sign structure carries it. Both
QTc slopes are **negative** -- rising sapanisertib concentration is associated
with QTc *shortening*, not prolongation -- while the RR slope is **positive**
(RR lengthening, i.e. heart-rate slowing). A user who assumes a common sign
convention across the three files will misread one of them.

# Population

```{r population, echo = FALSE}
pop <- readModelDb("Patel_2020_sapanisertib_QTcI")()$population
```

Forty-four adults with a radiographically or clinically evaluable advanced solid
tumor were enrolled at 5 United States centers (NCT02197572) and received the
single 40 mg dose. Median age was 59.5 years (range 22-79), 16 of 44 (36%) were
male, and median weight was 70.6 kg (range 49-122). Reported race was White in
30 (68%), Black in 9 (20%) and not reported in 5 (11%). Disease stage was IV or
higher in 34 of 44 (77%) and 42 of 44 (95%) had received prior antineoplastic
therapy; ECOG performance status was 0 in 13 (30%), 1 in 30 (68%) and 2 in
1 (2%) (Patel 2020 Table 1).

Two features of the cohort matter for interpreting the models. First, the
eligibility criteria **excluded** patients with baseline QTcF above 430 msec
(men) or 450 msec (women), congenital long QT syndrome, diabetes mellitus,
gastrointestinal disease, or significant active cardiovascular or pulmonary
disease (Patel 2020 Supplemental Methods, "Patients"), so the cohort is
pre-selected against QT-vulnerable patients. Second, **the analysis populations
differ**: all 44 patients contributed to the concentration-effect analysis these
model files encode, whereas the separate repeated-measures least-squares-mean
analysis (Patel 2020 Table 2 and Supplementary Table S1) used only the 32
patients with a complete day -1 / day 1 Holter series. The models below must not
be gated against Table 2 or Table S1.

The same information is available programmatically via each model's `population`
metadata, e.g. `readModelDb("Patel_2020_sapanisertib_QTcI")()$population`.

# Source trace

Every value in every `ini()` block comes from a regression-summary table printed
**inside** a figure panel. Those in-panel tables carry full printed-value
authority -- they are printed parameter tables, not digitised or curve-fitted
values -- but they are invisible to `pdftotext` and to docling because Figures 4
and S2 are raster images. They were recovered with
`pdftoppm -r 400 -f 10` on the article PDF (Figures 4A and 4C, journal page 885)
and by unzipping `cpdd808-sup-0001-suppmat.docx` and rendering
`word/media/image4.png` (Figure S2).

| Model | Parameter | Value | Source location (printed Estimate / SE / df / t / P / 95% bootstrap CI) |
|-------|-----------|-------|--------------------------------------------------------------------------|
| QTcI | `e0` | -0.304 msec | Fig 4A panel table, Intercept: -0.304 / 1.755 / 409 / -0.173 / 0.863 / (-3.757, 3.177) |
| QTcI | `slope` | -0.033 msec per ng/mL | Fig 4A panel table, Slope: -0.033 / 0.011 / 409 / -3.137 / 0.002 / (-0.060, -0.013) |
| QTcI | `addSd` | `fixed(0)` | Not reported in source -- typical-value model (see Errata) |
| QTcF | `e0` | -2.189 msec | Fig 4C panel table, Intercept: -2.189 / 1.137 / 440 / -1.926 / 0.055 / (-4.618, 0.220) |
| QTcF | `slope` | -0.016 msec per ng/mL | Fig 4C panel table, Slope: -0.016 / 0.006 / 440 / -2.446 / 0.015 / (-0.033, -0.003) |
| QTcF | `addSd` | `fixed(0)` | Not reported in source -- typical-value model (see Errata) |
| RR | `e0` | -25.504 msec | Fig S2 panel table, Intercept: -25.504 / 9.453 / 470 / -2.698 / 0.007 / (-44.154, -7.851) |
| RR | `slope` | +0.147 msec per ng/mL | Fig S2 panel table, Slope: 0.147 / 0.034 / 470 / 4.278 / 0.000 / (0.082, 0.242) |
| RR | `addSd` | `fixed(0)` | Not reported in source -- typical-value model (see Errata) |
| All | `CP_SAPANISERTIB_NGML` | ng/mL, time-varying | Patel 2020 Figures 4A / 4C / S2 x-axis, "Concentration (ng mL-1)" |
| All | Equation form | linear, intercept + slope | Patel 2020 Results, "Relationship of Sapanisertib Concentration and QTc" / "... and Heart Rate" |

## Verification of the printed panel tables

Because the six estimates were read off raster images, each panel table was
checked for internal arithmetic consistency before being trusted. The recomputed
two-sided P-value from each printed `(t, df)` pair must reproduce the printed
P-value, and each printed bootstrap confidence interval must contain its own
point estimate.

```{r panel-verification}
panels <- tibble::tribble(
  ~model,  ~parameter,   ~est,     ~se,    ~df, ~t_print, ~p_print, ~ci_lo,   ~ci_hi,
  "QTcI",  "Intercept",  -0.304,   1.755,  409, -0.173,   0.863,    -3.757,    3.177,
  "QTcI",  "Slope",      -0.033,   0.011,  409, -3.137,   0.002,    -0.060,   -0.013,
  "QTcF",  "Intercept",  -2.189,   1.137,  440, -1.926,   0.055,    -4.618,    0.220,
  "QTcF",  "Slope",      -0.016,   0.006,  440, -2.446,   0.015,    -0.033,   -0.003,
  "RR",    "Intercept", -25.504,   9.453,  470, -2.698,   0.007,   -44.154,   -7.851,
  "RR",    "Slope",       0.147,   0.034,  470,  4.278,   0.000,     0.082,    0.242
) |>
  dplyr::mutate(
    p_recomputed = 2 * stats::pt(-abs(t_print), df),
    p_agrees     = abs(p_recomputed - p_print) < 5e-4,
    ci_contains  = est >= ci_lo & est <= ci_hi
  )

panels |>
  dplyr::transmute(
    Model                 = model,
    Parameter             = parameter,
    Estimate              = est,
    SE                    = se,
    df                    = df,
    `t (printed)`         = t_print,
    `P (printed)`         = p_print,
    `P from (t, df)`      = signif(p_recomputed, 3),
    `P agrees`            = p_agrees,
    `CI contains estimate` = ci_contains
  ) |>
  knitr::kable(
    caption = paste(
      "Internal consistency of the six printed in-panel regression summaries.",
      "Every P-value reproduces from its own t-value and residual degrees of",
      "freedom, and every bootstrap CI contains its point estimate."
    )
  )

stopifnot(
  all(panels$p_agrees),
  all(panels$ci_contains)
)
```

# Independent corroboration of the two QTc lines

Table 4 of the paper tabulates model-predicted `Delta`QTcI and `Delta`QTcF at
three geometric-mean Cmax anchors -- 36.9 ng/mL (4 mg once daily), 235 ng/mL
(30 mg once weekly) and 297 ng/mL (the 40 mg dose of this study). Three points
**over-determine** a two-parameter line, so Table 4 can be inverted and compared
against the Figure 4 panel tables. This is a genuinely independent printed
source: it is set as text in the article body, not drawn inside a figure.

```{r table4-inversion}
table4 <- tibble::tribble(
  ~dose_mg, ~cmax_ngml, ~dQTcI,  ~dQTcF,
  4,          36.9,      -1.538,  -2.762,
  30,        235.0,      -8.159,  -5.837,
  40,        297.0,     -10.231,  -6.800
)

invert_line <- function(y) {
  fit <- stats::lm(y ~ table4$cmax_ngml)
  c(
    intercept = unname(stats::coef(fit)[1]),
    slope     = unname(stats::coef(fit)[2]),
    r_squared = summary(fit)$r.squared
  )
}

# Table 4 itself prints only three decimals, so the inversion inherits that
# granularity. Enumerating the 27 corner perturbations of +/-0.0005 on the
# three anchors bounds how far the inverted intercept and slope can move.
invert_band <- function(y) {
  corners <- expand.grid(a = c(-1, 0, 1), b = c(-1, 0, 1), c = c(-1, 0, 1))
  co <- t(apply(corners, 1, function(r) {
    stats::coef(stats::lm((y + 5e-4 * as.numeric(r)) ~ table4$cmax_ngml))
  }))
  c(
    intercept_lo = min(co[, 1]), intercept_hi = max(co[, 1]),
    slope_lo     = min(co[, 2]), slope_hi     = max(co[, 2])
  )
}

inverted <- dplyr::bind_cols(
  tibble::tibble(model = c("QTcI", "QTcF")),
  dplyr::bind_rows(
    as.data.frame(t(invert_line(table4$dQTcI))),
    as.data.frame(t(invert_line(table4$dQTcF)))
  ),
  dplyr::bind_rows(
    as.data.frame(t(invert_band(table4$dQTcI))),
    as.data.frame(t(invert_band(table4$dQTcF)))
  )
) |>
  dplyr::mutate(
    panel_intercept = c(-0.304, -2.189),
    panel_slope     = c(-0.033, -0.016),
    # Two printed sources, each rounded to three decimals, are consistent when
    # their rounding intervals overlap.
    intercept_consistent =
      intercept_lo - 5e-4 <= panel_intercept + 5e-4 &
      intercept_hi + 5e-4 >= panel_intercept - 5e-4,
    slope_consistent =
      slope_lo - 5e-4 <= panel_slope + 5e-4 &
      slope_hi + 5e-4 >= panel_slope - 5e-4
  )

inverted |>
  dplyr::transmute(
    Model                        = model,
    `Intercept (Table 4)`        = round(intercept, 5),
    `Intercept band from Table 4 rounding` =
      paste0("[", round(intercept_lo, 4), ", ", round(intercept_hi, 4), "]"),
    `Intercept (Fig 4 panel)`    = panel_intercept,
    `Slope (Table 4)`            = signif(slope, 6),
    `Slope (Fig 4 panel)`        = panel_slope,
    `R-squared of the 3 anchors` = signif(r_squared, 10)
  ) |>
  knitr::kable(
    caption = paste(
      "Inverting Table 4's three collinear Cmax anchors recovers the Figure 4",
      "panel estimates. Both sources print three decimals, so agreement is",
      "assessed by overlap of the two rounding intervals rather than by exact",
      "equality: the inverted QTcI intercept -0.30475 does not itself round to",
      "the printed -0.304, but Table 4's own rounding admits any intercept in",
      "[-0.3055, -0.3040], which overlaps the panel's [-0.3045, -0.3035]."
    )
  )

stopifnot(
  # The three Table 4 anchors are collinear to numerical precision, which is
  # what licenses the inversion at all.
  all(inverted$r_squared > 0.9999999),
  # The two independent printed sources are mutually consistent.
  all(inverted$intercept_consistent),
  all(inverted$slope_consistent),
  # And the point inversion sits within 0.001 of each printed panel value.
  all(abs(inverted$intercept - inverted$panel_intercept) < 1e-3),
  all(abs(inverted$slope - inverted$panel_slope) < 1e-3)
)
```

A third, sharper check falls out of this. `Estimate / SE` computed from the
printed three-decimal panel values does **not** reproduce the printed `t` for
either slope (`-0.033 / 0.011 = -3.04`, printed `-3.137`), which on its own could
look like a transcription error. It is not: dividing the *unrounded* slope
recovered from Table 4 by the printed `t` recovers a standard error that rounds
back to the printed SE. Both printed numbers are simply rounded to three
decimals, and the two independent sources agree once the rounding is undone.

```{r se-reconstruction}
se_check <- inverted |>
  dplyr::transmute(
    Model                  = model,
    `Unrounded slope`      = signif(slope, 6),
    `t printed in panel`   = c(-3.137, -2.446),
    `Implied SE`           = signif(slope / c(-3.137, -2.446), 5),
    `Implied SE rounded`   = round(slope / c(-3.137, -2.446), 3),
    `SE printed in panel`  = c(0.011, 0.006),
    `Naive t from printed values` = round(c(-0.033, -0.016) / c(0.011, 0.006), 3)
  )
knitr::kable(
  se_check,
  caption = paste(
    "The standard error implied by the Table-4-derived unrounded slope and the",
    "printed t-value rounds back to the printed SE, so the apparent",
    "Estimate/SE/t inconsistency in the panels is three-decimal rounding.",
    "This cross-checks the slope AND the SE against two independent printed",
    "sources."
  )
)

stopifnot(
  all(se_check$`Implied SE rounded` == se_check$`SE printed in panel`)
)
```

The `Delta`RR model has no counterpart to Table 4 -- Table 4 covers the two QTc
endpoints only -- so its verification rests on the internal
Estimate / SE / t / P / CI redundancy shown above. There, `Estimate / SE`
reproduces the printed `t` directly for the intercept
(`-25.504 / 9.453 = -2.698`, exactly as printed).

# Virtual cohort

The paper presents all three models in the **concentration domain** (Figures 4A,
4C and S2 plot the endpoint against sapanisertib concentration; Table 4 gives
point predictions at three concentrations). There is no time-domain model output
in the paper, so the primary reproduction below is a concentration grid rather
than a simulated time course.

The grid spans 0 to 1000 ng/mL, matching the x-axis extent of all three figure
panels; the observed concentration-effect dataset ranged from below the limit of
quantification to a maximum individual Cmax of 976 ng/mL (Patel 2020
Supplementary Table S2, Cmax median 317, range 94-976). Because the models are
purely algebraic with no time dependence, one observation event per grid point is
sufficient -- and well under the 200-per-arm cohort cap.

```{r cohort}
conc_grid <- seq(0, 1000, by = 10) # ng/mL; matches the Figure 4 / S2 x-axis

events <- tibble::tibble(
  conc = conc_grid
) |>
  dplyr::mutate(
    id                   = seq_len(dplyr::n()),
    CP_SAPANISERTIB_NGML = conc,
    time                 = 0,
    evid                 = 0, # observation event
    amt                  = 0,
    cmt                  = NA_character_
  )
```

# Simulation

Each model returns its change-from-time-matched-baseline endpoint directly as the
observation variable (`QTcI`, `QTcF`, `dRR`). No etas are declared -- the source
reports no variance components -- so `zeroRe()` is unnecessary and `omega = NA`
must not be passed.

```{r simulate}
mod_qtci <- readModelDb("Patel_2020_sapanisertib_QTcI")
mod_qtcf <- readModelDb("Patel_2020_sapanisertib_QTcF")
mod_rr   <- readModelDb("Patel_2020_sapanisertib_RR")

sim_qtci <- rxode2::rxSolve(mod_qtci, events = events, keep = "conc") |>
  as.data.frame()
sim_qtcf <- rxode2::rxSolve(mod_qtcf, events = events, keep = "conc") |>
  as.data.frame()
sim_rr <- rxode2::rxSolve(mod_rr, events = events, keep = "conc") |>
  as.data.frame()

cmax_anchors <- tibble::tibble(
  dose  = c("4 mg", "30 mg", "40 mg"),
  cmax  = c(36.9, 235.0, 297.0)
)
```

The three slopes must carry the signs the paper reports, and the two QTc
endpoints must remain far below the 20-msec threshold the paper treats as
clinically relevant for an oncology DQT study, across the whole observed
concentration range.

```{r sign-and-threshold-checks}
stopifnot(
  # Both QTc endpoints decrease with concentration (the paper's negative finding).
  all(diff(sim_qtci$QTcI) < 0),
  all(diff(sim_qtcf$QTcF) < 0),
  # The RR interval lengthens with concentration (heart-rate slowing).
  all(diff(sim_rr$dRR) > 0),
  # No QTc prolongation anywhere in the observed range, let alone 10 or 20 msec.
  max(sim_qtci$QTcI) <= 0,
  max(sim_qtcf$QTcF) <= 0
)
```

# Replicate published figures

## Figure 4A -- `Delta`QTcI versus sapanisertib concentration

```{r figure-4A}
ggplot(sim_qtci, aes(conc, QTcI)) +
  geom_hline(yintercept = c(10, 20), linetype = "dashed",
             colour = "#0072B2") +
  geom_vline(data = cmax_anchors, aes(xintercept = cmax),
             linetype = "dashed", colour = "#D55E00") +
  geom_line(linewidth = 1) +
  labs(
    x = "Sapanisertib plasma concentration (ng/mL)",
    y = "DeltaQTcI (msec)",
    title = "Figure 4A -- DeltaQTcI vs sapanisertib concentration",
    caption = paste(
      "Replicates the fitted line of Figure 4A of Patel 2020.",
      "Vertical lines: geometric-mean Cmax for 4, 30 and 40 mg.",
      "Horizontal lines: the 10 and 20 msec reference levels."
    )
  )
```

## Figure 4C -- `Delta`QTcF versus sapanisertib concentration

```{r figure-4C}
ggplot(sim_qtcf, aes(conc, QTcF)) +
  geom_hline(yintercept = c(10, 20), linetype = "dashed",
             colour = "#0072B2") +
  geom_vline(data = cmax_anchors, aes(xintercept = cmax),
             linetype = "dashed", colour = "#D55E00") +
  geom_line(linewidth = 1) +
  labs(
    x = "Sapanisertib plasma concentration (ng/mL)",
    y = "DeltaQTcF (msec)",
    title = "Figure 4C -- DeltaQTcF vs sapanisertib concentration",
    caption = paste(
      "Replicates the fitted line of Figure 4C of Patel 2020.",
      "The QTcF slope is roughly half the magnitude of the QTcI slope."
    )
  )
```

## Supplementary Figure S2 -- `Delta`RR versus sapanisertib concentration

```{r figure-S2}
zero_cross <- -readModelDb("Patel_2020_sapanisertib_RR")()$theta[["e0"]] /
  readModelDb("Patel_2020_sapanisertib_RR")()$theta[["slope"]]

ggplot(sim_rr, aes(conc, dRR)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey40") +
  geom_vline(data = cmax_anchors, aes(xintercept = cmax),
             linetype = "dashed", colour = "#D55E00") +
  geom_line(linewidth = 1) +
  labs(
    x = "Sapanisertib plasma concentration (ng/mL)",
    y = "DeltaRR (msec)",
    title = "Supplementary Figure S2 -- DeltaRR vs sapanisertib concentration",
    caption = paste0(
      "Replicates the fitted line of Figure S2 of Patel 2020. ",
      "The line crosses zero at ", round(zero_cross, 1), " ng/mL: below that ",
      "concentration the model predicts RR shortening (faster heart rate) ",
      "and above it RR lengthening."
    )
  )

stopifnot(
  # Zero crossing is -e0 / slope = 25.504 / 0.147.
  abs(zero_cross - 25.504 / 0.147) < 1e-9,
  abs(zero_cross - 173.5) < 0.1
)
```

# Validation gate: reproducing Table 4

Table 4 is a **zero-parameter gate** on the two QTc models: it contains six
model-predicted values that the packaged models must reproduce with no fitting,
no seed and no free choices at all. The models ship the *printed* three-decimal
panel estimates, so that every number in `ini()` is a number printed in the
paper; that rounding is the only source of disagreement.

```{r table4-gate}
predict_endpoint <- function(mod, conc) {
  ev <- tibble::tibble(
    id                   = seq_along(conc),
    CP_SAPANISERTIB_NGML = conc,
    time                 = 0,
    evid                 = 0,
    amt                  = 0,
    cmt                  = NA_character_
  )
  as.data.frame(rxode2::rxSolve(mod, events = ev))
}

gate <- table4 |>
  dplyr::mutate(
    model_dQTcI = predict_endpoint(mod_qtci, cmax_ngml)$QTcI,
    model_dQTcF = predict_endpoint(mod_qtcf, cmax_ngml)$QTcF,
    diff_dQTcI  = model_dQTcI - dQTcI,
    diff_dQTcF  = model_dQTcF - dQTcF
  )

gate |>
  dplyr::transmute(
    `Dose`                       = paste(dose_mg, "mg"),
    `Geometric mean Cmax (ng/mL)` = cmax_ngml,
    `Paper DeltaQTcI (msec)`     = dQTcI,
    `Model DeltaQTcI (msec)`     = round(model_dQTcI, 3),
    `QTcI diff`                  = round(diff_dQTcI, 3),
    `Paper DeltaQTcF (msec)`     = dQTcF,
    `Model DeltaQTcF (msec)`     = round(model_dQTcF, 3),
    `QTcF diff`                  = round(diff_dQTcF, 3)
  ) |>
  knitr::kable(
    caption = paste(
      "Zero-parameter validation against Patel 2020 Table 4. Model values use",
      "the printed three-decimal Figure 4 panel estimates; all six agree with",
      "the paper to within 0.15 msec, which is entirely slope rounding."
    )
  )

stopifnot(
  # Deterministic comparison -- no random draws on either side -- so a tight
  # bound on every one of the six values is the correct assertion.
  all(abs(gate$diff_dQTcI) < 0.15),
  all(abs(gate$diff_dQTcF) < 0.15)
)
```

Using the unrounded slopes and intercepts recovered from Table 4 instead of the
printed three-decimal values collapses the residual to numerical noise, which
confirms that the 0.15 msec offset above is rounding and nothing else. Users who
need to reproduce Table 4 digit-for-digit should override `ini()` with the
unrounded values.

```{r table4-gate-unrounded}
exact <- gate |>
  dplyr::mutate(
    exact_dQTcI = inverted$intercept[1] + inverted$slope[1] * cmax_ngml,
    exact_dQTcF = inverted$intercept[2] + inverted$slope[2] * cmax_ngml,
    ediff_dQTcI = exact_dQTcI - dQTcI,
    ediff_dQTcF = exact_dQTcF - dQTcF
  )

exact |>
  dplyr::transmute(
    `Dose`                        = paste(dose_mg, "mg"),
    `Paper DeltaQTcI (msec)`      = dQTcI,
    `Unrounded model DeltaQTcI`   = round(exact_dQTcI, 5),
    `QTcI diff`                   = signif(ediff_dQTcI, 3),
    `Paper DeltaQTcF (msec)`      = dQTcF,
    `Unrounded model DeltaQTcF`   = round(exact_dQTcF, 5),
    `QTcF diff`                   = signif(ediff_dQTcF, 3)
  ) |>
  knitr::kable(
    caption = paste(
      "The same gate using the unrounded intercept and slope back-solved from",
      "Table 4. Residuals fall to below 0.001 msec, isolating the rounding."
    )
  )

stopifnot(
  all(abs(exact$ediff_dQTcI) < 2e-3),
  all(abs(exact$ediff_dQTcF) < 2e-3)
)
```

# Illustrative time course

The three packaged models are algebraic in concentration; driving them over time
requires an external sapanisertib concentration trajectory, which this paper does
not supply as a model. The chunk below builds a purely **illustrative**
one-compartment oral trajectory in order to demonstrate the time-varying-covariate
usage pattern. It is anchored on three numbers printed in Supplementary Table S2
for the study's 40 mg dose -- geometric mean Cmax 297 ng/mL, median Tmax 1.52 h
and geometric mean AUC-infinity 2480 ng*h/mL -- which exactly determine the three
Bateman parameters.

**This reconstruction is not part of any model file and is not a fit reported by
the paper.** It also carries a diagnostic message: matching Cmax, Tmax and
AUC-infinity simultaneously forces an apparent terminal half-life of about 4.6 h,
roughly half the 9.06 h geometric-mean terminal half-life the paper reports.
A one-compartment form cannot satisfy both the observed Cmax/AUC ratio and the
observed terminal slope, which is direct evidence of multi-compartment
disposition that the paper's non-compartmental analysis leaves unmodelled. Anyone
needing a quantitative sapanisertib PK driver should obtain a genuine popPK model
rather than use this illustration.

```{r illustrative-pk}
dose_mg <- 40
cl_f    <- dose_mg * 1000 / 2480 # L/h, from AUCinf = Dose / (CL/F)
tmax_h  <- 1.52
cmax    <- 297

# For a one-compartment oral model, Cmax = (Dose / V) * exp(-kel * Tmax), which
# with CL/F = V * kel determines V, then kel, then ka via the Tmax identity.
v_f <- stats::uniroot(
  function(v) v - (dose_mg * 1000 / cmax) * exp(-tmax_h * cl_f / v),
  interval = c(100, 1000)
)$root
kel <- cl_f / v_f
ka <- stats::uniroot(
  function(ka) log(ka / kel) / (ka - kel) - tmax_h,
  interval = c(kel * 1.0001, 500)
)$root

bateman <- function(t) {
  dose_mg * ka / (v_f * (ka - kel)) * (exp(-kel * t) - exp(-ka * t)) * 1000
}

c(
  `CL/F (L/h)`               = round(cl_f, 2),
  `V/F (L)`                  = round(v_f, 1),
  `ka (1/h)`                 = round(ka, 3),
  `implied t1/2 (h)`         = round(log(2) / kel, 2),
  `reproduced Cmax (ng/mL)`  = round(bateman(tmax_h), 1),
  `reproduced AUCinf`        = round(dose_mg / cl_f * 1000, 0)
)

stopifnot(
  # The reconstruction hits its three printed anchors by construction; this
  # asserts the numerics converged, not anything about the paper.
  abs(bateman(tmax_h) - cmax) < 0.01,
  abs(dose_mg / cl_f * 1000 - 2480) < 1
)
```

```{r illustrative-timecourse}
tc_events <- tibble::tibble(
  time = c(seq(0, 12, by = 0.1), seq(12.5, 48, by = 0.5))
) |>
  dplyr::mutate(
    id                   = 1L,
    CP_SAPANISERTIB_NGML = bateman(time),
    evid                 = 0,
    amt                  = 0,
    cmt                  = NA_character_
  )

tc <- dplyr::bind_rows(
  as.data.frame(rxode2::rxSolve(mod_qtci, events = tc_events)) |>
    dplyr::transmute(time, value = QTcI, endpoint = "DeltaQTcI"),
  as.data.frame(rxode2::rxSolve(mod_qtcf, events = tc_events)) |>
    dplyr::transmute(time, value = QTcF, endpoint = "DeltaQTcF"),
  as.data.frame(rxode2::rxSolve(mod_rr, events = tc_events)) |>
    dplyr::transmute(time, value = dRR, endpoint = "DeltaRR")
)

ggplot(tc, aes(time, value)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey40") +
  geom_line(linewidth = 0.9) +
  facet_wrap(~endpoint, ncol = 1, scales = "free_y") +
  labs(
    x = "Time after the single 40 mg dose (h)",
    y = "Change from time-matched baseline (msec)",
    title = "Illustrative time course of the three concentration-effect models",
    caption = paste(
      "NOT a paper figure. The sapanisertib trajectory is an illustrative",
      "one-compartment reconstruction anchored on the printed Cmax, Tmax and",
      "AUCinf; see the caveats above."
    )
  )

peak_conc_time <- tc_events$time[which.max(tc_events$CP_SAPANISERTIB_NGML)]

stopifnot(
  # The two QTc endpoints reach their most negative value at the same TIME as
  # the concentration peak, and the RR endpoint its most positive: the models
  # are direct-effect, with no hysteresis (the paper checked for hysteresis
  # graphically and found none -- Methods, "Statistical Analyses"). Compared on
  # time rather than row index so the assertion does not depend on the row
  # ordering rxSolve returns.
  with(tc[tc$endpoint == "DeltaQTcI", ], time[which.min(value)]) ==
    peak_conc_time,
  with(tc[tc$endpoint == "DeltaQTcF", ], time[which.min(value)]) ==
    peak_conc_time,
  with(tc[tc$endpoint == "DeltaRR", ], time[which.max(value)]) ==
    peak_conc_time
)
```

# Reconciling the RR model with the reported heart-rate time course

The concentration-`Delta`RR fit and the paper's heart-rate time course point in
apparently opposite directions, and it is worth being explicit that they do not
conflict.

Supplementary Table S1 reports the least-squares-mean change in heart rate
rising steadily with time, from -4.4 bpm at 0.5 h to a maximum of +12.4 bpm at
24 h after dosing. But at 24 h the sapanisertib concentration has fallen far
below Cmax -- roughly 11 ng/mL on the illustrative trajectory above, given the
reported 9 h half-life. So the observed net tachycardia occurs when drug
concentrations are *low*, and it is carried in the concentration-effect model by
the large negative intercept (-25.5 msec of RR, i.e. a faster heart rate than
time-matched baseline at zero concentration), not by the slope. The positive
slope acts in the opposite direction, partially offsetting that intercept as
concentration rises.

```{r rr-reconciliation}
rr_points <- tibble::tibble(
  scenario = c("Zero concentration", "4 mg Cmax", "30 mg Cmax", "40 mg Cmax"),
  conc     = c(0, 36.9, 235.0, 297.0)
) |>
  dplyr::mutate(
    dRR = predict_endpoint(mod_rr, conc)$dRR
  )

rr_points |>
  dplyr::transmute(
    Scenario                       = scenario,
    `Sapanisertib conc (ng/mL)`    = conc,
    `Model DeltaRR (msec)`         = round(dRR, 2)
  ) |>
  knitr::kable(
    caption = paste(
      "Model-predicted DeltaRR at zero concentration and at the three",
      "geometric-mean Cmax anchors. The intercept carries an RR shortening",
      "(faster heart rate); the positive slope offsets it as concentration",
      "rises, turning DeltaRR positive above about 174 ng/mL."
    )
  )

stopifnot(
  # The intercept-only prediction is an RR shortening, and the 40 mg Cmax
  # prediction is an RR lengthening: the sign flip within the studied range is
  # the whole reconciliation, so it is asserted rather than only narrated.
  rr_points$dRR[rr_points$scenario == "Zero concentration"] < 0,
  rr_points$dRR[rr_points$scenario == "40 mg Cmax"] > 0
)
```

Note also that the two analyses use different populations (44 patients for the
concentration-effect fits, 32 for the least-squares-mean time course) and
different scales (RR interval in msec versus heart rate in bpm, which are
reciprocally related), so no quantitative gate between them is appropriate.

# Assumptions and deviations

- **All six parameter values come from regression tables printed inside raster
  figure panels.** Figures 4A and 4C of the article and Figure S2 of the
  supplement each embed a `Parameter | Estimate | SE | df | t-value | P-value |
  95% Bootstrap CI` table inside the plot area. These are printed parameter
  tables and carry full printed-value authority -- nothing here was digitised off
  a curve or fitted -- but they are invisible to every text-extraction path.
  Recovery required `pdftoppm -r 400 -f 10` on the article PDF and unzipping
  `cpdd808-sup-0001-suppmat.docx` to render `word/media/image4.png`. Reviewers
  re-checking these values must render the images; grepping the text will find
  nothing. The verification tables above exist precisely so the transcription can
  be audited without re-reading the images.

- **Inter-individual variability is not quantified in the source.** The Methods
  state that "Mixed-effects models were subsequently developed to describe the
  direct effect of sapanisertib on change from time-matched baseline in RR, QTcI,
  and QTcF", and the residual degrees of freedom (409, 440, 470 against 44
  patients) are consistent with a per-subject random intercept. But no variance
  estimate, random-effect standard deviation or shrinkage value is reported for
  any of the three models. Per the standing operator policy on unreported IIV,
  the packaged models omit the eta declarations and ship typical-value-only
  encodings. The only surviving traces of the variance components are Table 4's
  threshold-exceedance probability columns and the Figure 4B / 4D 90% prediction
  intervals; neither inverts to a unique variance, so no value is guessed. Users
  who need an IIV envelope must impose one externally.

- **Residual error is not reported numerically.** `addSd <- fixed(0)` in all
  three files, per the standing operator policy, so each model returns its
  deterministic typical-value prediction. This follows the same encoding as
  `Darpo_2014_racSotalol_QTcI.R` / `_QTcF.R`.

- **Table 4's exceedance probabilities are deliberately not gated.** The paper
  reports the estimated probability of `Delta`QTc exceeding 30 and 60 msec at
  each Cmax anchor (at most 0.0139 and 0.0001 respectively). Reproducing those
  numbers requires the unreported variance components, and back-solving them from
  the probabilities does not identify a unique variance: the implied total
  standard deviation is not constant across the three anchors (about 9.7, 14.8
  and 18.3 msec at 4, 30 and 40 mg for `Delta`QTcI), so the reported
  probabilities evidently also carry concentration uncertainty, parameter
  uncertainty, or both. Rather than fit a variance to match a probability column,
  the gap is recorded here.

- **No structural PK model exists in the source.** Sapanisertib PK was
  characterised by non-compartmental analysis only (geometric mean Cmax 297
  ng/mL, AUC-infinity 2480 ng*h/mL, CL/F 16.1 L/h, terminal half-life 9.06 h
  geometric mean; Supplementary Table S2), and no sapanisertib popPK model exists
  in the nlmixr2lib registry as of this extraction. All three models therefore
  consume `CP_SAPANISERTIB_NGML` (ng/mL) as a time-varying covariate. The
  illustrative reconstruction above is a vignette device only, and its 4.6 h
  implied half-life versus the reported 9.06 h shows a one-compartment form is
  inadequate for this drug.

- **Three model files, one vignette.** The paper fits the three
  concentration-effect models independently, one per endpoint, with different
  residual degrees of freedom -- they are not sub-models of a joint fit. Per the
  standing replicate-the-author's-structure policy that maps to three `.R` files
  and, because they come from one paper, this single vignette.

- **The RR endpoint is named `dRR`, the two QTc endpoints are not `d`-prefixed.**
  This asymmetry within the file set is deliberate and was an operator ruling
  (sidecar `oare_PMC7586797` request-001 q1, answered 2026-09-02). The register
  had no RR entry of either form. `dRR` was ratified as a canonical sibling of
  the existing `dHR` change-from-baseline heart-rate endpoint, on the
  scale-incomparability grounds that entry already states -- an absolute RR
  interval is roughly 850 msec whereas this `Delta`RR spans -25 to +18 msec -- and
  because the competing bare-name precedent (`QTc`, `QTcF`, `QTcI`) is for
  absolute interval names. The two QTc siblings keep their bare corrected-interval
  names in line with the existing registry usage in
  `Darpo_2014_racSotalol_QTcF.R`, `Mukker_2026_tuvusertib_QTcF.R` and
  `Zhou_2025_fruquintinib_*`; the per-model `units` field documents the
  change-from-baseline semantic in each case.

- **Reported race percentages do not sum to 100.** Table 1 lists White 68%,
  Black 20% and not reported 11%, summing to 99% from independent rounding of
  30, 9 and 5 of 44. The `population` metadata records the printed percentages
  as given.

- **PKNCA validation is not applicable.** PKNCA is the validation surface for
  drug-concentration endpoints; all three endpoints here are derived
  electrocardiographic PD quantities in msec, not plasma concentrations, and the
  paper reports no concentration-time model to run NCA against. The Table 4
  zero-parameter gate is the appropriate validation for a linear mixed-effects PD
  model whose entire content is its coefficient table, and it is supplemented
  above by the Table-4 inversion cross-check, the panel-table internal-consistency
  check, and the sign and threshold assertions.
