---
title: "Eteplirsen (Patel 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Eteplirsen (Patel 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
mdl <- rxode2::rxode(readModelDb("Patel_2025_eteplirsen"))
cat("* Citation: ", mdl$reference, "\n", sep = "")
cat("* Description: ", mdl$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1002/psp4.70001>
* Supplement (Method S1, Figures S1-S6, Tables S1-S2): Data S1 of the
  Supporting Information, retrieved as `PSP4-14-891-s001.pdf`. Table S1
  (ethnicity, race and exon-deletion mutations) and Table S2 (simulated
  steady-state exposures and the weight-by-age distribution) are both used
  below.

Eteplirsen is a phosphorodiamidate morpholino oligomer (PMO) approved by the US
FDA for patients with Duchenne muscular dystrophy (DMD) carrying a mutation
amenable to exon 51 skipping. Patel 2025 is the first published population PK
analysis for the antisense-oligonucleotide class in DMD. It pools six clinical
studies to characterise disposition across the full target age range and to
support extension of the uniform 30 mg/kg weekly dosing regimen to patients as
young as 6 months.

# Population

The analysis dataset held 3258 quantifiable eteplirsen plasma concentrations
from 157 male patients with exon-51-skip-amenable DMD, pooled from six studies
(Patel 2025 Table 1): NCT03218995 (4658-102, ages 6-48 months), NCT01396239
(4658-201), NCT01540409 (4658-202), NCT02420379 (4658-203), NCT00844597
(4658-28) and NCT02255552 (4658-301). Median age was 8.37 years (range 0.55 to
16.4), median body weight 27.1 kg (range 6.80 to 68.9) and median eGFR
145 mL/min/1.73 m^2 (range 85.5 to 180). All patients are male because DMD is
X-linked recessive. Race was 84% White, 6% Asian, 2% Black or African American,
1% Pacific Islander, 3% Other and 4% missing; 8% were Hispanic or Latino
(Table S1). Doses ranged from 0.5 to 50 mg/kg weekly as an approximately 1-hour
IV infusion; 37 samples (1.1%) below the limit of quantification were excluded.

Because creatinine is a by-product of the muscle breakdown that characterises
DMD, creatinine clearance is not a valid renal marker in this population. The
authors therefore estimated eGFR from serum cystatin C using the CKD-EPI
equation (Methods section 2.3). No subject had an eGFR below
85.5 mL/min/1.73 m^2, so the model carries no information about renal
impairment.

The same information is available programmatically via the model's `population`
metadata:

```{r population}
str(readModelDb("Patel_2025_eteplirsen")()$population, max.level = 1)
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Patel_2025_eteplirsen.R` carries an in-file comment
naming its source location. They are collected here for review.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl_agegt4` (CL, age > 4 y) | 6.98 L/h | Table 2, theta1 (95% CI 6.68-7.31) |
| `lcl_agele4` (CL, age <= 4 y) | 4.97 L/h | Table 2, theta2 (95% CI 4.48-5.47) |
| `lvc` (V1) | 2.21 L | Table 2, theta3 (95% CI 1.93-2.55) |
| `lvp` (V2) | 2.23 L | Table 2, theta4 (95% CI 1.90-2.64) |
| `lq` (Q2) | 0.334 L/h | Table 2, theta5 (95% CI 0.262-0.422) |
| `lvp2` (V3) | 4.12 L | Table 2, theta6 (95% CI 3.81-4.43) |
| `lq2` (Q3) | 3.49 L/h | Table 2, theta7 (95% CI 3.29-3.70) |
| `e_crcl_cl` (eGFR on CL) | 1.60 | Table 2, theta8 (95% CI 1.33-1.92) |
| `e_wt_cl`, `e_wt_q`, `e_wt_q2` | 0.75 (fixed) | Eq. 1, 4, 6; Results 3.3 |
| `e_wt_vc`, `e_wt_vp`, `e_wt_vp2` | 1 (fixed) | Eq. 2, 3, 5; Results 3.3 |
| Reference weight | 37 kg | Results 3.4 narrative |
| Reference eGFR | 145 mL/min/1.73 m^2 | Results 3.4 narrative; Table 1 median |
| Age cutoff | 4 years | Eq. 1 `ifelse(Age > 4 yrs, theta1, theta2)` |
| `etalcl` variance | 0.05375 = log(1 + 0.235^2) | Table 2 omega_CL 23.5 CV%; Methods 2.5.1 %CV formula |
| `etalvc` variance | 0.22154 = log(1 + 0.498^2) | Table 2 omega_V1 49.8 CV% |
| `etalvp` variance | 0.40348 = log(1 + 0.705^2) | Table 2 omega_V2 70.5 CV% |
| `etalq` variance | 0.72314 = log(1 + 1.030^2) | Table 2 omega_Q2 103 CV% |
| IIV covariances | 0.0541, 0.110, 0.167, 0.193, 0.295, 0.470 | Table 2 Cov() rows |
| `expSd` | 0.34086 = sqrt(log(1 + 0.351^2)) | Table 2 sigma_RV 35.1 CV%; Methods 2.5.1 |
| Structure: 3 compartments, linear elimination | n/a | Results 3.3; Eq. 1-6 |
| Zero-order input (1-h IV infusion) | n/a | Methods 2.2; Results 3.3 |
| Residual error additive on log domain | n/a | Methods 2.5.1 |

The two clearance thetas are the only stratum-specific parameters. Everything
else -- volumes, inter-compartmental clearances, the eGFR exponent, the whole
4x4 IIV block and the residual error -- is shared between the two age strata in
a single joint NONMEM fit, so the model is one file rather than two.

Table 2 reports the structural thetas already back-transformed (its Units column
reads L/h and L) while Methods section 2.5.1 states that parameters were
"modeled in the log-domain" with `P_i = exp(P_hat + eta_Pi)`. Each `ini()` entry
is therefore written `log(<Table 2 estimate>)`. The same section prints the
transform used for the variance terms, `%CV = 100 * sqrt(exp(omega^2) - 1)`,
which is what fixes the diagonal of the IIV block as `log(1 + CV^2)` rather than
`CV^2`. The resulting 4x4 matrix is positive definite (smallest eigenvalue
0.0119, correlations 0.50 to 0.87).

# Virtual cohort

Original observed data are not publicly available. The simulations below use
virtual populations whose covariate distributions approximate the published
demographics. Body weights are drawn from log-normal distributions matched to
the median and 5th/95th percentiles of Table S2, which lists the weight
distribution the authors used for their own stochastic simulation. Age is drawn
uniformly within each age band and enters the model only through the 4-year
cutoff. eGFR is held at the reference value of 145 mL/min/1.73 m^2 -- the paper
reports the weight-by-age sources for its virtual population but does not state
how eGFR was simulated (see Assumptions).

```{r cohort}
set.seed(20250203)

N_PER_ARM <- 200L

age_groups <- tibble::tribble(
  ~agegrp,       ~age_lo, ~age_hi, ~wt_med, ~wt_lo, ~wt_hi,
  "0.5 to <2 y",     0.5,     2.0,    10.2,   8.19,   13.7,
  "2 to <4 y",       2.0,     4.0,    14.6,   11.3,   18.2,
  "4 to <7 y",       4.0,     7.0,    19.6,   14.7,   27.9,
  "7 to <=16 y",     7.0,    16.0,    38.1,   22.3,   74.3
)

# Dense sampling over the first day (where all the structure is) and a coarse
# tail out to the end of the 168 h dosing interval.
OBS_TIMES <- c(seq(0, 24, by = 0.1), seq(26, 168, by = 2))

make_cohort <- function(n, age_lo, age_hi, wt_med, wt_lo, wt_hi, label,
                        mgkg = 30, crcl = 145, id_offset = 0L,
                        obs_times = OBS_TIMES) {
  sdlog <- (log(wt_hi) - log(wt_lo)) / (2 * stats::qnorm(0.95))
  subj <- tibble::tibble(
    id    = id_offset + seq_len(n),
    AGE   = stats::runif(n, age_lo, age_hi),
    WT    = stats::rlnorm(n, log(wt_med), sdlog),
    CRCL  = crcl,
    mgkg  = mgkg,
    agegrp = label
  )
  doses <- subj |>
    dplyr::mutate(
      time = 0, evid = 1L, cmt = "central",
      amt  = mgkg * WT,
      # rate = amt / 1 h reproduces the approximately 1-hour IV infusion of
      # Methods section 2.2 (the "zero-order absorption" of Results 3.3).
      rate = mgkg * WT
    )
  obs <- tidyr::expand_grid(subj, time = obs_times) |>
    dplyr::mutate(evid = 0L, cmt = "central", amt = NA_real_, rate = NA_real_)
  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  lapply(seq_len(nrow(age_groups)), function(i) {
    g <- age_groups[i, ]
    make_cohort(
      n = N_PER_ARM, age_lo = g$age_lo, age_hi = g$age_hi,
      wt_med = g$wt_med, wt_lo = g$wt_lo, wt_hi = g$wt_hi,
      label = g$agegrp, id_offset = (i - 1L) * N_PER_ARM
    )
  })
)

# Disjoint IDs across cohorts are mandatory: duplicate IDs silently merge into
# one subject that receives the summed dose.
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
stopifnot(dplyr::n_distinct(events$id) == N_PER_ARM * nrow(age_groups))
```

# Simulation

```{r simulate}
mod <- readModelDb("Patel_2025_eteplirsen")

sim <- rxode2::rxSolve(
  mod, events = events,
  keep = c("agegrp", "WT", "AGE", "CRCL", "mgkg")
) |>
  as.data.frame()

# rxSolve can silently drop subjects; assert the count survived.
stopifnot(dplyr::n_distinct(sim$id) == N_PER_ARM * nrow(age_groups))
```

Weekly dosing with a short half-life produces essentially no accumulation, which
is why Patel 2025 could assume steady state at the first simulated dose
(Methods section 2.8). The check below confirms that concentration at the end of
the 168 h dosing interval is a negligible fraction of Cmax, so a single-dose
profile over 0-168 h *is* the steady-state dosing interval.

```{r accumulation-check}
sim |>
  dplyr::group_by(id) |>
  dplyr::summarise(
    cmax  = max(Cc),
    ctau  = Cc[which.max(time)],
    .groups = "drop"
  ) |>
  dplyr::summarise(
    `Max C(168 h) / Cmax across subjects` = format(max(ctau / cmax), scientific = TRUE, digits = 3),
    `Median C(168 h) / Cmax`              = format(stats::median(ctau / cmax), scientific = TRUE, digits = 3)
  ) |>
  knitr::kable(caption = "Accumulation check: residual concentration at the end of the weekly dosing interval, relative to Cmax.")
```

# Replicate published figures

## Figure 1 -- dose-normalised concentrations superimpose across dose levels

Patel 2025 Figure 1 plots dose-normalised eteplirsen concentration versus time
coloured by dose group and reports that the profiles "were superimposed without
evidence of any clear trends", supporting linear PK across 0.5-50 mg/kg. The
model is linear by construction, so this is a check that the packaged
implementation carries no dose-dependent term.

```{r figure-1}
dose_levels <- c(0.5, 2, 10, 30, 50)

events_dose <- dplyr::bind_rows(
  lapply(seq_along(dose_levels), function(i) {
    make_cohort(
      n = 50L, age_lo = 0.55, age_hi = 16.4,
      wt_med = 27.1, wt_lo = 8.0, wt_hi = 62.0,
      label = paste0(dose_levels[i], " mg/kg"),
      mgkg = dose_levels[i], id_offset = (i - 1L) * 50L,
      obs_times = seq(0, 24, by = 0.25)
    )
  })
)
stopifnot(!anyDuplicated(unique(events_dose[, c("id", "time", "evid")])))

sim_dose <- rxode2::rxSolve(
  mod, events = events_dose, keep = c("agegrp", "WT", "mgkg")
) |>
  as.data.frame() |>
  dplyr::rename(dose_group = agegrp)

sim_dose |>
  dplyr::filter(Cc > 0) |>
  dplyr::mutate(cc_norm = Cc / mgkg) |>
  dplyr::group_by(dose_group, time) |>
  dplyr::summarise(Q50 = stats::median(cc_norm), .groups = "drop") |>
  ggplot(aes(time, Q50, colour = dose_group)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(
    x = "Time after start of infusion (h)",
    y = "Dose-normalised Cc (ug/mL per mg/kg)",
    colour = "Dose group",
    title = "Figure 1 -- dose-normalised concentration-time profiles",
    caption = "Replicates Figure 1 of Patel 2025."
  )
```

## Figure 3 -- concentration versus time after dose

Patel 2025 Figure 3a is a prediction-corrected VPC of eteplirsen concentration
versus time after dose. The panel below shows the simulated median with the
5th-95th prediction interval over the first 24 h, the window the paper's own
simulation used.

```{r figure-3}
sim |>
  dplyr::filter(time > 0, time <= 24, Cc > 0) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    Q05 = stats::quantile(Cc, 0.05),
    Q50 = stats::median(Cc),
    Q95 = stats::quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(
    x = "Time after start of infusion (h)", y = "Cc (ug/mL)",
    title = "Figure 3 -- simulated median and 5th-95th prediction interval",
    caption = "Replicates the shape of Figure 3a of Patel 2025 (30 mg/kg weekly, all age groups pooled)."
  )
```

The profile shows the features the paper describes in Results section 3.2:
concentration peaks at the end of the 1-hour infusion, a rapid distribution
phase follows over the next 2-3 h, and a slower elimination phase runs out over
the remainder of the day.

## Figure 4 -- covariate effects on normalised steady-state exposure

Figure 4 of Patel 2025 is a forest plot of steady-state AUC and Cmax relative to
a typical subject, evaluated **at a reference dose** (so the body-weight effect
is not cancelled by weight-based dosing), against a reference range of
0.8-1.67. The reference subject is a patient older than 4 years weighing 37 kg
with an eGFR of 145 mL/min/1.73 m^2.

```{r figure-4}
mod_typical <- rxode2::zeroRe(mod)

REF_DOSE <- 30 * 37  # mg, the 30 mg/kg dose of the 37 kg reference patient

typical_exposure <- function(wt = 37, crcl = 145, age = 8) {
  ev <- tibble::tibble(
    id = 1L, time = c(0, seq(0, 168, by = 0.05)),
    evid = c(1L, rep(0L, length(seq(0, 168, by = 0.05)))),
    cmt = "central",
    amt = c(REF_DOSE, rep(NA_real_, length(seq(0, 168, by = 0.05)))),
    rate = c(REF_DOSE, rep(NA_real_, length(seq(0, 168, by = 0.05)))),
    WT = wt, CRCL = crcl, AGE = age
  ) |>
    dplyr::arrange(time, dplyr::desc(evid))
  s <- rxode2::rxSolve(mod_typical, events = ev) |> as.data.frame()
  s <- s[order(s$time), ]
  c(
    cmax = max(s$Cc),
    auc  = sum(diff(s$time) * (utils::head(s$Cc, -1) + utils::tail(s$Cc, -1)) / 2)
  )
}

ref <- typical_exposure()

forest <- dplyr::bind_rows(
  tibble::tibble(
    covariate = "Body weight (kg)",
    level     = c("6.80 (min)", "68.9 (max)"),
    value     = c(6.80, 68.9)
  ) |>
    dplyr::rowwise() |>
    dplyr::mutate(e = list(typical_exposure(wt = value))) |>
    dplyr::ungroup(),
  tibble::tibble(
    covariate = "eGFR (mL/min/1.73 m^2)",
    level     = c("85.5 (min)", "180 (max)"),
    value     = c(85.5, 180)
  ) |>
    dplyr::rowwise() |>
    dplyr::mutate(e = list(typical_exposure(crcl = value))) |>
    dplyr::ungroup(),
  tibble::tibble(
    covariate = "Age stratum",
    level     = "<= 4 years",
    value     = 2
  ) |>
    dplyr::rowwise() |>
    dplyr::mutate(e = list(typical_exposure(age = value))) |>
    dplyr::ungroup()
) |>
  dplyr::mutate(
    `AUCss ratio`   = vapply(e, function(x) unname(x["auc"]), numeric(1)) / ref[["auc"]],
    `Cmax,ss ratio` = vapply(e, function(x) unname(x["cmax"]), numeric(1)) / ref[["cmax"]]
  ) |>
  dplyr::select(-e, -value)

forest |>
  tidyr::pivot_longer(c(`AUCss ratio`, `Cmax,ss ratio`),
                      names_to = "metric", values_to = "ratio") |>
  ggplot(aes(ratio, paste(covariate, level, sep = ": "))) +
  annotate("rect", xmin = 0.8, xmax = 1.67, ymin = -Inf, ymax = Inf,
           fill = "grey80", alpha = 0.5) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_point(size = 2.5) +
  facet_wrap(~metric) +
  labs(
    x = "Exposure relative to the typical subject", y = NULL,
    title = "Figure 4 -- covariate effects at a fixed reference dose",
    caption = "Replicates Figure 4 of Patel 2025. Grey band is the 0.8-1.67 reference range."
  )
```

```{r figure-4-table}
forest |>
  knitr::kable(
    digits = 3,
    caption = "Normalised steady-state exposure by covariate level, at a fixed 1110 mg dose."
  )
```

This reproduces the paper's three qualitative conclusions (Results section 3.5):

* **Body weight is the only significant covariate.** Both the minimum and the
  maximum observed weight put the point estimate outside the 0.8-1.67 band, for
  both AUC and Cmax.
* **Age <= 4 years is not significant.** The AUC ratio is
  `r round(forest$"AUCss ratio"[forest$covariate == "Age stratum"], 2)`, inside
  the band. Propagating the published 95% CIs of the two clearance thetas gives
  a worst case of 7.31 / 4.48 = `r round(7.31 / 4.48, 2)`, still inside the
  band -- matching the paper's statement that "the point estimate and 95% CI for
  the exposure parameters were within reference range".
* **eGFR is retained but not significant.** Note that the min/max eGFR values
  used above *do* fall outside the band, whereas the paper reports the low-eGFR
  effect as non-significant and the high-eGFR effect only as "inconclusive".
  With the point estimate of 1.60, the fixed-dose exposure ratio
  `(eGFR/145)^-1.60` crosses the upper bound of 1.67 at eGFR = 105 and the lower
  bound of 0.8 at eGFR = 167. Both crossing points lie inside the observed
  85.5-180 range, so Figure 4 of the paper must have used a narrower covariate
  range than the Table 1 min-max -- most likely observed percentiles. The paper
  does not report which values Figure 4 used, so this is noted rather than
  reconstructed.

## Figure 5 and Table S2 -- steady-state exposure by age group

```{r figure-5, fig.height = 4}
exposure <- sim |>
  dplyr::group_by(agegrp, id) |>
  dplyr::arrange(time, .by_group = TRUE) |>
  dplyr::summarise(
    `Cmax,ss (ug/mL)`  = max(Cc),
    `AUCss (ug*h/mL)`  = sum(diff(time) * (utils::head(Cc, -1) + utils::tail(Cc, -1)) / 2),
    .groups = "drop"
  ) |>
  dplyr::mutate(agegrp = factor(agegrp, levels = age_groups$agegrp))

exposure |>
  tidyr::pivot_longer(c(`Cmax,ss (ug/mL)`, `AUCss (ug*h/mL)`),
                      names_to = "metric", values_to = "value") |>
  ggplot(aes(agegrp, value)) +
  geom_boxplot(outlier.size = 0.6) +
  facet_wrap(~metric, scales = "free_y") +
  labs(
    x = "Age group", y = NULL,
    title = "Figure 5 -- steady-state exposure at 30 mg/kg weekly by age group",
    caption = "Replicates Figure 5 of Patel 2025."
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

## Structural identity: AUCss x CL = Dose, per subject

For any linear model the steady-state exposure over a full dosing interval
satisfies `AUCss = Dose / CL` exactly. `rxSolve` returns each subject's
individual clearance, so this identity can be asserted per subject rather than
compared as a median -- a much sharper test, and the one that pins down where
the Table S2 discrepancy discussed below actually lives.

```{r auc-identity}
identity_check <- sim |>
  dplyr::group_by(id, agegrp) |>
  dplyr::arrange(time, .by_group = TRUE) |>
  dplyr::summarise(
    cl    = dplyr::first(cl),
    WT    = dplyr::first(WT),
    auc   = sum(diff(time) * (utils::head(Cc, -1) + utils::tail(Cc, -1)) / 2),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    dose     = 30 * WT,
    implied  = auc * cl,
    rel_err  = implied / dose - 1
  )

# Trapezoidal error on the coarse post-24 h grid is the only source of
# deviation; assert it is under 1% for every single subject.
stopifnot(max(abs(identity_check$rel_err)) < 0.01)

tibble::tibble(
  `Max |AUC * CL / Dose - 1| across all 800 subjects` =
    sprintf("%.4f%%", 100 * max(abs(identity_check$rel_err)))
) |>
  knitr::kable(caption = "Per-subject structural identity AUCss x CL = Dose.")
```

Because the identity holds to four decimal places for every subject, any
mismatch against a published AUCss value is a statement about the published
clearance, not about the simulation.

# PKNCA validation

NCA is computed with PKNCA over the full 168 h steady-state dosing interval,
grouped by age band so the results line up with Table S2.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, agegrp)

# Guarantee a time-zero record per subject (pre-infusion Cc is 0 for an IV
# infusion). Without it PKNCA warns "Requesting an AUC range starting (0)
# before the first measurement" once per subject.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, agegrp) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, agegrp, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

dose_df <- events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, agegrp)

conc_obj <- PKNCA::PKNCAconc(
  sim_nca, Cc ~ time | agegrp + id,
  concu = "ug/mL", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  dose_df, amt ~ time | agegrp + id,
  doseu = "mg"
)

intervals <- data.frame(
  start    = 0,
  end      = 168,
  cmax     = TRUE,
  tmax     = TRUE,
  auclast  = TRUE,
  cav      = TRUE,
  half.life = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)
```

```{r pknca-summary}
as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "half.life")) |>
  dplyr::group_by(agegrp, PPTESTCD) |>
  dplyr::summarise(
    median = stats::median(PPORRES),
    p2.5   = stats::quantile(PPORRES, 0.025),
    p97.5  = stats::quantile(PPORRES, 0.975),
    .groups = "drop"
  ) |>
  dplyr::mutate(agegrp = factor(agegrp, levels = age_groups$agegrp)) |>
  dplyr::arrange(agegrp, PPTESTCD) |>
  dplyr::rename(
    "Age group"  = agegrp,
    "Parameter"  = PPTESTCD,
    "Median"     = median,
    "2.5th pctl" = p2.5,
    "97.5th pctl" = p97.5
  ) |>
  knitr::kable(digits = 2, caption = "Simulated steady-state NCA by age group (n = 200 per group).")
```

## Comparison against published NCA

Table S2 of Patel 2025 reports simulated steady-state AUC and Cmax by age group
as median (95% CI). Those are the reference values below.

```{r nca-compare}
published <- tibble::tribble(
  ~agegrp,        ~cmax,  ~auclast,
  "0.5 to <2 y",  119.0,  159.0,
  "2 to <4 y",    112.0,  135.0,
  "4 to <7 y",     98.1,  106.0,
  "7 to <=16 y",  119.0,  140.0
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by        = "agegrp",
  params    = c("cmax", "auclast"),
  units     = c(cmax = "ug/mL", auclast = "ug*h/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated vs. published steady-state exposure (Patel 2025 Table S2). * differs from reference by more than 20%."
)
```

```{r nca-diffs, echo = FALSE}
# Numeric per-band differences, joined BY KEY (never by row position), so the
# narrative below cannot drift from the table above.
sim_med <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "auclast")) |>
  dplyr::group_by(agegrp, PPTESTCD) |>
  dplyr::summarise(sim = stats::median(PPORRES), .groups = "drop")

diffs <- published |>
  tidyr::pivot_longer(c(cmax, auclast),
                      names_to = "PPTESTCD", values_to = "ref") |>
  dplyr::left_join(sim_med, by = c("agegrp", "PPTESTCD")) |>
  dplyr::mutate(pct = 100 * (sim - ref) / ref)
stopifnot(!any(is.na(diffs$sim)))

pctRange <- function(which_param) {
  p <- diffs$pct[diffs$PPTESTCD == which_param]
  sprintf("%+.1f%% to %+.1f%%", min(p), max(p))
}
```

The **Cmax,ss** column reproduces across all four age bands
(`r pctRange("cmax")`), with no row flagged. The dispersion reproduces too: the
simulated 2.5th-97.5th percentile intervals in the NCA summary table above sit
close to the published 95% CIs of Table S2 (for the youngest group, 80.0-167
published). Since Cmax is driven mainly by the three volumes and the infusion
duration, this validates the structural model, V1/V2/V3, and the IIV block.

The **AUCss** column differs by `r pctRange("auclast")`: it reproduces for the
youngest band alone and is over-predicted for the other three. Given the
per-subject `AUCss x CL = Dose` identity asserted above, this is a statement
about the published table rather than about the implementation -- see Errata
below.

# Assumptions and deviations

* **eGFR distribution.** Patel 2025 states the sources for the weight-by-age
  distribution of its virtual population (CDC NHANES for age < 2 years, West
  et al. 2013 for ages 2-12, the analysis dataset above 12) but does not say how
  eGFR was simulated. eGFR is therefore held at the reference value of
  145 mL/min/1.73 m^2 -- the cohort median from Table 1 and the normalising
  constant in Eq. 1 -- for every simulated subject. Because the eGFR exponent is
  1.60, exposure is sensitive to this choice: a 10% shift in eGFR moves AUC by
  about 16%.
* **Age distribution.** Age is drawn uniformly within each published age band.
  Age enters the model only through the 4-year cutoff, so only the fraction of
  each band above or below 4 years matters, and every band lies wholly on one
  side of the cutoff.
* **Weight distribution.** Body weight is drawn from a log-normal fitted to the
  median and 5th/95th percentiles reported in Table S2. The underlying
  distribution shape is not published.
* **Steady state via a single dose.** Patel 2025 assumed steady state at the
  first simulated dose. The accumulation check above confirms this is exact to
  within a negligible residual, so the vignette simulates a single dose over one
  168 h dosing interval rather than a dosing series.
* **Figure 4 covariate range.** The covariate values used in the paper's forest
  plot are not reported numerically; the min/max of Table 1 are used here, and
  the range that would reproduce the paper's eGFR conclusions is reconstructed
  in the narrative rather than assumed.
* **No non-paper-derived parameter values.** Every `ini()` entry comes from
  Patel 2025 Table 2 or Equations 1-6. Nothing was digitised from a figure,
  obtained by correspondence, or carried from an upstream model, and no value
  was tuned to improve agreement with any published output.

# Errata and unresolved discrepancies

**1. Table S2 AUCss is not reproducible from Equations 1-6.** For any linear
model the steady-state exposure over a dosing interval satisfies the exact
identity `AUCss = Dose / CL`. With a weight-based dose of 30 mg/kg and Eq. 1,
this gives

```
AUCss = 30 * WT / (CL_typ * (WT/37)^0.75 * (eGFR/145)^1.60)
      proportional to WT^0.25   (at fixed CL stratum and eGFR)
```

so within one age stratum AUCss must **increase** with body weight. Table S2
reports AUCss *falling* from 159 to 135 ug*h/mL as the median weight rises from
10.2 to 14.6 kg -- both rows in the age <= 4 y stratum, where the typical
clearance is the same 4.97 L/h. No value of the unreported eGFR distribution
can produce that ordering together with the published allometric exponent.

Three candidate explanations were tested and all fail:

* **An unreported eGFR distribution.** Solving for the eGFR that would
  reproduce each published AUCss gives 147, 172, 169 and 158 mL/min/1.73 m^2
  for the four bands. Those are implausible as age-band medians -- the study
  cohorts closest to each band have observed median eGFR of 142 (ages
  6-48 months), 149 (ages 4-6 years) and 143 (ages 7-16 years) per Table 1 --
  and imposing them degrades the Cmax,ss agreement into a systematic negative
  bias, whereas at eGFR = 145 the Cmax residuals are scattered about zero.
* **A different reference weight.** No single alternative reconciles the four
  rows: the implied reference weights are 36, 26, 27 and 31 kg.
* **A 24-hour integration window.** Methods section 2.8 states that simulated
  subjects were "followed 24 h post dose", which raises the possibility that
  Table S2 reports AUC over 0-24 h rather than over the full 168 h dosing
  interval. It does not: at the four Table S2 median weights, typical-value
  AUC(0-24 h) recovers 99.8% or more of AUC(0-168 h), so the two windows are
  interchangeable here and the choice cannot account for a 30% gap.

The Cmax,ss column of Table S2 does agree in magnitude, which is why the
discrepancy is isolated to the AUCss column: Cmax is roughly three times less
sensitive to clearance than AUC is. It is worth noting, though, that the Cmax
column shows the *same* within-stratum ordering anomaly in miniature (119 then
112 ug/mL as weight rises from 10.2 to 14.6 kg, where the model requires a
slight increase); the ordering is simply small enough to stay inside the 20%
tolerance. Both columns therefore point the same way.

None of this affects the packaged model, whose parameters are taken verbatim
from Table 2 and Equations 1-6 and which satisfies `AUCss = Dose / CL` exactly
for every simulated subject.

**2. The eGFR sensitivity quoted in Results section 3.4 is internally
inconsistent with Eq. 1.** The paper states that raising eGFR to
160 mL/min/1.73 m^2 increases CL by 14.5% and that lowering it to 130 decreases
CL by 18.4%. Applying Eq. 1 with theta8 = 1.60 and the stated reference of 145
gives +17.1% and -16.0% respectively. Both quoted percentages are consistent
with a reference eGFR near 147 rather than 145; the difference is immaterial
(about 1.4% in the normalising constant) and the model uses the reference of 145
that Eq. 1 and the Results 3.4 narrative both state explicitly.

**3. Epsilon shrinkage is high.** Table 2 notes epsilon-shrinkage of 67.7% for
the residual variability, so individual-level residual diagnostics from this
model should be treated with caution. Eta shrinkage is modest (11.0%, 22.1%,
24.9% and 20.4% for CL, V1, V2 and Q2).

**4. Model applicability outside the observed eGFR range.** No subject had an
eGFR below 85.5 mL/min/1.73 m^2. The paper states explicitly that "model
application below observed eGFR range would be limited". With an exponent of
1.60 on a strongly supranormal reference of 145, extrapolation into renal
impairment will over-predict the reduction in clearance.
