---
title: "Cariprazine, desmethyl-cariprazine and didesmethyl-cariprazine (Periclou 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cariprazine, desmethyl-cariprazine and didesmethyl-cariprazine (Periclou 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Periclou_2021_cariprazine"))$reference`
* Description: `r rxode2::rxode(readModelDb("Periclou_2021_cariprazine"))$description`
* Article: [Eur J Drug Metab Pharmacokinet 2021;46(1):53-69](https://doi.org/10.1007/s13318-020-00650-4) (open access)
* Supplement: Electronic Supplementary Material 1 of the article (Supplemental Equation Sets 1-3 and Supplemental Tables 1-4)

Periclou 2021 reports two sets of sequentially fitted parent-metabolite
models, and both are packaged:

* `Periclou_2021_cariprazine` -- the **final models on the updated dataset**
  (Table 2, Supplemental Equation Sets 1 and 3). This is the paper's headline
  model and the one to use for simulation.
* `Periclou_2021_cariprazine_initial` -- the final models on the **initial
  dataset** (Supplemental Equation Set 2, Supplemental Table 3), fitted before
  the rich-sampling Japanese study A002-A11 was available and restricted to
  samples collected within 25 h of a dose. It is superseded by the updated
  model and is included for completeness.

The cariprazine, DCAR and DDCAR models were estimated one after another with
the preceding analyte's individual parameters fixed, and all cariprazine
eliminated was assumed to become DCAR and all DCAR eliminated to become DDCAR.
The DCAR and DDCAR parameters are therefore apparent values, and each file
couples the three analytes into a single ODE system.

# Population

The updated model was developed on 2199 adults (18-65 years, mean 39.2 years)
from three phase 1 studies, eight phase 2/3 studies and the Japanese open-label
study A002-A11, of whom 66% were men. Body weight averaged 78.9 kg (SD 18.7,
range 33.1-155.1). The cohort was 45.6% White, 34.9% Black, 14.3% Asian (mainly
patients at Indian sites), 1.7% Japanese and 3.5% other race (Supplemental
Table 2). Patients had schizophrenia or manic/mixed episodes of bipolar I
disorder and received oral cariprazine once daily at 0.5-12.5 mg; data after
doses of 15 mg/day or more were excluded. The long-term open-label studies
RGH-MD-11 and RGH-MD-17 were held out for external validation. Creatinine
clearance and CYP2D6 metaboliser status had no effect on exposure.

The same information is available programmatically via
`readModelDb("Periclou_2021_cariprazine")()$population`.

# Model structure

Supplemental Equation Set 1 and Figure 1 of the paper define nine states:

* `depot` receives the dose as a **zero-order input** over `DUR` = 2.57 h and
  empties into `central` by first-order absorption (`Ka`).
* Cariprazine distributes from `central` to two peripheral compartments
  (`peripheral1` through `Q3/F`, `peripheral2` through `Q4/F`).
* Cariprazine elimination (`CL/Vc`) is the formation rate of DCAR in
  `central_dcar`, which exchanges with `peripheral1_dcar`.
* DCAR elimination (`DCL/DVc`) feeds a single delay compartment, `transit1`,
  which empties into `central_ddcar` at `DDKtr` = 0.0269 1/h. The delay reflects
  that only 2.6% of patients had measurable DDCAR before day 2.
* DDCAR exchanges with `peripheral1_ddcar` and is eliminated by `DDCL/DDVc`.

Because the phase 2/3 studies did not sample after the first dose, the authors
added a **first-dose shift** `FD` (1 for concentrations after the first dose, 0
otherwise) on cariprazine `Vc/F`, `VP1/F` and `Q3/F` and on DCAR `Vc/F` and
`Vp/F`. The packaged model builds `FD` internally from `dosenum()`, so no data
column is needed: it equals 1 until the second dose is given.

Dose records must use `rate = -2` so that rxode2 applies `dur(depot)`; a dose
without it enters the depot as a bolus.

```{r model}
mod <- readModelDb("Periclou_2021_cariprazine")
mod_typ <- rxode2::zeroRe(rxode2::rxode2(mod))

# Molar-to-mass factors (ng/mL per nM) implied by the paper's own tables:
# Supplemental Table 4 (ng/mL) divided by Table 3 (nM) gives 295 / 690.2,
# 78.0 / 188.7 and 629 / 1573.8, i.e. molecular weights of 427.4, 413.4 and
# 399.7 g/mol for cariprazine, DCAR and DDCAR.
mw <- c(CAR = 0.4274, DCAR = 0.4134, DDCAR = 0.3997)

# Event-table builder. Observation rows sit on the ODE state `central` and
# nominate endpoint 1 through dvid; every observation row returns Cc, Cc_dcar
# and Cc_ddcar together.
make_events <- function(id, dose_times, amts, obs_times, covs) {
  ev <- rbind(
    data.frame(id = id, time = dose_times, amt = amts, rate = -2, evid = 1L,
               cmt = "depot", dvid = NA_integer_),
    data.frame(id = id, time = obs_times, amt = NA_real_, rate = NA_real_,
               evid = 0L, cmt = "central", dvid = 1L)
  )
  ev <- ev[order(ev$time, -ev$evid), ]
  cbind(ev, covs[rep(1, nrow(ev)), , drop = FALSE], row.names = NULL)
}

to_nM <- function(sim) {
  # rxSolve() omits the id column when only one subject is solved.
  if (!"id" %in% names(sim)) sim$id <- 1L
  sim |>
    dplyr::transmute(
      id, time,
      CAR = Cc / mw[["CAR"]],
      DCAR = Cc_dcar / mw[["DCAR"]],
      DDCAR = Cc_ddcar / mw[["DDCAR"]],
      Total = CAR + DCAR + DDCAR
    )
}

trap <- function(t, y) sum(diff(t) * (head(y, -1) + tail(y, -1)) / 2)

male_white <- function(WT) {
  data.frame(WT = WT, SEXF = 0, RACE_BLACK = 0, RACE_ASIAN = 0, RACE_JAPANESE = 0)
}
```

# Source trace

Every `ini()` value carries an in-file comment naming its source. They are
collected here.

| Parameter | Value | Source location |
|-----------|-------|-----------------|
| `ld1` (DUR) | log(2.57 h) | Table 2 |
| `lka` (Ka) | log(0.352 1/h) | Table 2 |
| `lcl` (CL/F) | log(21.5 L/h) | Table 2; Suppl Eq 10 |
| `e_wt_cl`, `e_black_cl`, `e_asian_cl`, `e_japanese_cl` | 0.0946, -0.0907, -0.178, -0.111 | Table 2; Suppl Eq 10 |
| `lvc` (Vc/F) | log(266 L) | Table 2; Suppl Eq 11 |
| `e_wt_vc`, `e_fd_vc` | 1.66, 2.84 (fixed) | Table 2; Suppl Eq 11 |
| `lq` (Q3/F), `e_fd_q` | log(0.431 L/h) fixed, 39.4 fixed | Table 2; Suppl Eq 12 |
| `lvp` (VP1/F), `e_fd_vp` | log(149 L) fixed, 2.61 fixed | Table 2; Suppl Eq 13 |
| `lq2` (Q4/F), `lvp2` (VP2/F) | log(100 L/h), log(501 L), both fixed | Table 2 |
| `lcl_dcar` (DCL/F) | log(77.3 L/h) | Table 2; Suppl Eq 14 |
| `e_wt_cl_dcar`, `e_black_cl_dcar`, `e_asian_cl_dcar`, `e_japanese_cl_dcar`, `e_sexf_cl_dcar` | 0.578, 0.249, -0.0861, -0.145, -0.160 | Table 2; Suppl Eq 14 |
| `lvc_dcar` (DVc/F), `e_wt_vc_dcar`, `e_fd_vc_dcar` | log(128 L), 1.18, 1.27 fixed | Table 2; Suppl Eq 15 |
| `lq_dcar` (DQ/F) | log(78.5 L/h) | Table 2 |
| `lvp_dcar` (DVp/F), `e_fd_vp_dcar` | log(347 L), 0.535 fixed | Table 2; Suppl Eq 16 |
| `lcl_ddcar` (DDCL/F) | log(9.24 L/h) | Table 2; Suppl Eq 17 |
| `e_wt_cl_ddcar`, `e_black_cl_ddcar`, `e_asian_cl_ddcar`, `e_japanese_cl_ddcar` | 0.427, 0.547, -0.194, -0.156 | Table 2; Suppl Eq 17 |
| `lvc_ddcar` (DDVc/F) | log(1310 L) | Table 2; Suppl Eq 18 |
| `e_wt_vc_ddcar`, `e_black_vc_ddcar`, `e_asian_vc_ddcar`, `e_japanese_vc_ddcar` | 0.881, 0.676, -0.240, 0.0888 | Table 2; Suppl Eq 18 |
| `lq_ddcar` (DDQ/F), `lvp_ddcar` (DDVp/F) | log(0.386 L/h), log(258 L), both fixed | Table 2 |
| `lktr` (DDKtr) | log(0.0269 1/h) fixed | Table 2 |
| `etalka`, `etalcl`, `etalvc` | 0.8723, 0.0998, 0.7830 | Table 2 IIV 118%, 32.4%, 109% CV; `log(1 + CV^2)` |
| `etalcl_dcar`, `etalvc_dcar` | 0.1653, 0.8426 | Table 2 IIV 42.4%, 115% CV |
| `etalcl_ddcar`, `etalvc_ddcar` | 0.2848, 0.4733 | Table 2 IIV 57.4%, 77.8% CV |
| `propSd`, `propSd_dcar`, `propSd_ddcar` | 0 (fixed) | not reported (Table 2 footnote f) |
| ODEs `d/dt(depot)` ... `d/dt(peripheral1_ddcar)` | n/a | Suppl Equation Set 1, Eqs 1-9; Figure 1 |
| `dur(depot)` zero-order input | n/a | Suppl Eq 1 ('initial condition of Dose/DUR'); Figure 1 'R0 = Dose/DUR' |
| First-dose indicator `fd` | n/a | Suppl Equation Set 3 definition of FD; Section 3.2.2 |

The initial-dataset model (`Periclou_2021_cariprazine_initial`) takes its
values from Supplemental Table 3 and its covariate equations from Supplemental
Equations A1-A6, line by line as documented in that file.

# Typical-patient profiles (replicates Figure 5)

The paper's Section 3.6 describes Figure 5 as a simulation of an "overall
typical patient (79 kg, Caucasian/white adult male)". Solving the model at
79 kg puts cariprazine about 8.5% above every Figure 5a point from 2 to 12 h,
whereas at **84 kg** -- the typical Caucasian male the paper uses in Figures 6
and 7 -- the same points agree to within 3% for both cariprazine and DCAR.
Weight scales cariprazine `Vc/F` with exponent 1.66, so that is exactly the
size of offset a 5 kg difference produces. Figure 5 is therefore replicated at
84 kg. The figure values below were read from the published figure by the
maintainers (about +/-0.1 nM in panel a, +/-1 nM in panel b and +/-10 nM*h in
panel c).

```{r fig5-digitised}
fig5a <- tibble::tribble(
  ~analyte, ~time, ~conc,
  "CAR", 1, 0.72, "CAR", 2, 2.55, "CAR", 3, 4.82, "CAR", 4, 6.32,
  "CAR", 6, 7.50, "CAR", 8, 7.56, "CAR", 12, 6.87, "CAR", 24, 2.24,
  "DCAR", 2, 0.10, "DCAR", 3, 0.28, "DCAR", 4, 0.51, "DCAR", 6, 0.88,
  "DCAR", 8, 1.11, "DCAR", 12, 1.31, "DCAR", 24, 1.12,
  "DDCAR", 12, 0.05, "DDCAR", 24, 0.48
)
fig5b <- tibble::tribble(
  ~analyte, ~day, ~conc,
  "CAR", 1.5, 14.1, "CAR", 2, 10.0, "CAR", 3, 5.9, "CAR", 4, 3.3, "CAR", 5, 2.0,
  "CAR", 6, 1.3,
  "DDCAR", 2, 57.1, "DDCAR", 3, 54.9, "DDCAR", 4, 50.8, "DDCAR", 5, 45.7,
  "DDCAR", 6, 40.4, "DDCAR", 8, 31.4, "DDCAR", 10, 23.8, "DDCAR", 12, 17.8,
  "DDCAR", 14, 13.4, "DDCAR", 21, 5.0, "DDCAR", 28, 2.8
)
fig5c <- tibble::tribble(
  ~analyte, ~day, ~auc,
  "CAR", 7, 627, "CAR", 14, 655, "CAR", 84, 655,
  "DCAR", 7, 224, "DCAR", 14, 232, "DCAR", 84, 232,
  "DDCAR", 7, 471, "DDCAR", 14, 1078, "DDCAR", 21, 1290, "DDCAR", 28, 1370,
  "DDCAR", 84, 1430,
  "Total", 7, 1186, "Total", 14, 1825, "Total", 21, 2045, "Total", 28, 2117,
  "Total", 84, 2185
)
```

## Figure 5a: first 6 mg dose

```{r fig5a, fig.height = 4}
ev_sd <- make_events(1L, 0, 6, c(0, seq(0.25, 24, by = 0.25)), male_white(84))
sd84 <- rxode2::rxSolve(mod_typ, ev_sd, returnType = "data.frame") |> to_nM()

sd84 |>
  tidyr::pivot_longer(c(CAR, DCAR, DDCAR, Total), names_to = "analyte", values_to = "conc") |>
  ggplot(aes(time, conc, colour = analyte)) +
  geom_line() +
  geom_point(data = fig5a, shape = 1, size = 2) +
  labs(x = "Time after first dose (h)", y = "Concentration (nM)", colour = NULL,
       title = "Figure 5a: single 6 mg dose, typical 84 kg White male",
       caption = "Lines: packaged model. Open circles: values read from Figure 5a of Periclou 2021.")
```

```{r fig5a-check}
chk5a <- fig5a |>
  dplyr::rowwise() |>
  dplyr::mutate(model = sd84[[analyte]][which.min(abs(sd84$time - time))]) |>
  dplyr::ungroup() |>
  dplyr::mutate(pct_diff = 100 * (model - conc) / conc)
knitr::kable(chk5a, digits = 2, caption = "Model vs Figure 5a (nM).")

# Deterministic typical-value solve, so tight bounds are appropriate. Measured
# at authoring: cariprazine within 3% at 1-12 h, DCAR within 3% at 4-12 h.
# The 24 h points are a documented deviation (see Assumptions and deviations)
# and are excluded from the gate, as is DCAR at 2-3 h where the figure values
# (0.10 and 0.28 nM) sit at the resolution of the digitisation.
g_car <- chk5a$analyte == "CAR" & chk5a$time <= 12
g_dcar <- chk5a$analyte == "DCAR" & chk5a$time >= 4 & chk5a$time <= 12
stopifnot(
  sum(g_car) == 7, sum(g_dcar) == 4,
  max(abs(chk5a$pct_diff[g_car])) < 8,
  max(abs(chk5a$pct_diff[g_dcar])) < 8
)
```

## Figure 5b: washout after the last dose at steady state

```{r fig5b, fig.height = 4}
ev_ss <- make_events(1L, seq(0, 83 * 24, by = 24), 6,
                     c(0, seq(83 * 24, 111 * 24, by = 0.5)), male_white(84))
ss84 <- rxode2::rxSolve(mod_typ, ev_ss, returnType = "data.frame") |>
  to_nM() |>
  dplyr::filter(time >= 83 * 24) |>
  dplyr::mutate(day = (time - 83 * 24) / 24)

ss84 |>
  tidyr::pivot_longer(c(CAR, DCAR, DDCAR, Total), names_to = "analyte", values_to = "conc") |>
  ggplot(aes(day, conc, colour = analyte)) +
  geom_line() +
  geom_point(data = fig5b, shape = 1, size = 2) +
  labs(x = "Time after the last of 84 daily 6 mg doses (days)", y = "Concentration (nM)",
       colour = NULL, title = "Figure 5b: washout from steady state, typical 84 kg White male",
       caption = "Lines: packaged model. Open circles: values read from Figure 5b of Periclou 2021.")
```

```{r fig5b-check}
chk5b <- fig5b |>
  dplyr::rowwise() |>
  dplyr::mutate(model = ss84[[analyte]][which.min(abs(ss84$day - day))]) |>
  dplyr::ungroup() |>
  dplyr::mutate(pct_diff = 100 * (model - conc) / conc)
knitr::kable(chk5b, digits = 2, caption = "Model vs Figure 5b (nM).")

# Cariprazine washout reproduces within about 12% (measured +6% to +12% at
# authoring). DDCAR runs a near-constant 13-16% high over the whole washout
# (26% at day 21, where the figure value is small), the same offset seen in
# its steady-state level (see Figure 5c below), so its shape is reproduced
# and only its level differs.
stopifnot(
  max(abs(chk5b$pct_diff[chk5b$analyte == "CAR"])) < 15,
  abs(median(chk5b$pct_diff[chk5b$analyte == "DDCAR"])) < 25
)
```

## Figure 5c: daily AUC during titration to 6 mg

Figure 5c titrates 1.5, 3 and 4.5 mg on days 1-3 and 6 mg daily thereafter.

```{r fig5c, fig.height = 4}
amts_tit <- c(1.5, 3, 4.5, rep(6, 81))
ev_tit <- make_events(1L, seq(0, 83 * 24, by = 24), amts_tit,
                      seq(0, 84 * 24, by = 0.25), male_white(84))
tit84 <- rxode2::rxSolve(mod_typ, ev_tit, returnType = "data.frame") |> to_nM()

daily_auc <- tit84 |>
  dplyr::mutate(day = pmin(floor(time / 24) + 1, 84)) |>
  dplyr::group_by(day) |>
  dplyr::group_modify(function(d, key) {
    w <- tit84[tit84$time >= (key$day - 1) * 24 & tit84$time <= key$day * 24, ]
    data.frame(CAR = trap(w$time, w$CAR), DCAR = trap(w$time, w$DCAR),
               DDCAR = trap(w$time, w$DDCAR), Total = trap(w$time, w$Total))
  }) |>
  dplyr::ungroup() |>
  tidyr::pivot_longer(-day, names_to = "analyte", values_to = "auc")

ggplot(daily_auc, aes(day, auc, colour = analyte)) +
  geom_line() +
  geom_point(data = fig5c, shape = 1, size = 2) +
  labs(x = "Day of dosing", y = "AUC0-24 (nM*h)", colour = NULL,
       title = "Figure 5c: daily AUC0-24 during titration to 6 mg/day",
       caption = "Lines: packaged model. Open circles: values read from Figure 5c of Periclou 2021.")
```

```{r fig5c-check}
chk5c <- fig5c |>
  dplyr::left_join(daily_auc, by = c("analyte", "day"), suffix = c("_fig", "_model")) |>
  dplyr::mutate(pct_diff = 100 * (auc_model - auc_fig) / auc_fig)
knitr::kable(chk5c, digits = 1, caption = "Model vs Figure 5c (nM*h).")

# Figure 5c is not internally consistent: its Total curve sits about 130 nM*h
# BELOW the sum of its own CAR + DCAR + DDCAR curves at every day, so only the
# cariprazine curve is gated. At authoring the cariprazine plateau matched
# within 1%.
fig_sum_gap <- with(fig5c, auc[analyte == "Total" & day == 84] -
  sum(auc[analyte %in% c("CAR", "DCAR", "DDCAR") & day == 84]))
stopifnot(
  fig_sum_gap < -100,
  max(abs(chk5c$pct_diff[chk5c$analyte == "CAR"])) < 5
)
```

# Deterministic mass-balance check

The chain assumes that every molecule of cariprazine eliminated becomes DCAR
and every molecule of DCAR eliminated becomes DDCAR, and the paper's equations
carry amounts in dose units with no molecular-weight term. At steady state each
analyte's dosing-interval AUC must therefore equal `Dose / CL` of that analyte
(in ng*h/mL). This holds for any correct encoding and breaks at once if a
formation or elimination term is mis-wired.

```{r mass-balance}
ev_mb <- make_events(1L, seq(0, 139 * 24, by = 24), 6,
                     seq(139 * 24, 140 * 24, by = 0.05), male_white(79))
mb <- rxode2::rxSolve(mod_typ, ev_mb, returnType = "data.frame")
mb_tab <- data.frame(
  analyte = c("CAR", "DCAR", "DDCAR"),
  auc_tau = c(trap(mb$time, mb$Cc), trap(mb$time, mb$Cc_dcar), trap(mb$time, mb$Cc_ddcar)),
  dose_over_cl = 6000 / c(21.5, 77.3, 9.24)
) |>
  dplyr::mutate(pct_diff = 100 * (auc_tau - dose_over_cl) / dose_over_cl)
knitr::kable(mb_tab, digits = 2,
             caption = "Steady-state AUC over 24 h vs Dose/CL (ng*h/mL), typical 79 kg White male, day 140.")
stopifnot(max(abs(mb_tab$pct_diff)) < 0.5)
```

# Japanese patients (Supplemental Table 4)

Supplemental Table 4 gives the population-model mean steady-state exposure for
the 37 Japanese patients of study A002-A11 at 6 mg/day. Their mean weight is
not reported; the paper's typical Japanese male in Figure 7 weighs 78 kg.

```{r japanese}
jp <- data.frame(WT = 78, SEXF = 0, RACE_BLACK = 0, RACE_ASIAN = 0, RACE_JAPANESE = 1)
jp_sim <- rxode2::rxSolve(mod_typ, make_events(1L, seq(0, 139 * 24, by = 24), 6,
                                               seq(139 * 24, 140 * 24, by = 0.1), jp),
                          returnType = "data.frame")
jp_tab <- data.frame(
  analyte = c("CAR", "DCAR", "DDCAR"),
  model_auc = c(trap(jp_sim$time, jp_sim$Cc), trap(jp_sim$time, jp_sim$Cc_dcar),
                trap(jp_sim$time, jp_sim$Cc_ddcar)),
  suppl_table4_mean_auc = c(306, 105, 784)
) |>
  dplyr::mutate(pct_diff = 100 * (model_auc - suppl_table4_mean_auc) / suppl_table4_mean_auc)
knitr::kable(jp_tab, digits = 1,
             caption = "Typical Japanese male (78 kg) vs Supplemental Table 4 popPK mean AUC0-24 (ng*h/mL).")

# Cariprazine and DDCAR carry no sex effect, so the typical male should sit
# near the Japanese mean (measured +2.7% and -1.4%). DCAR clearance is 16%
# lower in women, and the Japanese cohort was not all male, so the DCAR row
# is shown but not gated.
stopifnot(max(abs(jp_tab$pct_diff[jp_tab$analyte != "DCAR"])) < 10)
```

# Covariate effects on total exposure (Figure 4)

The paper reports that every covariate-related change in steady-state Total
CAR exposure was within 36% of its comparator (White or Other for race, male for
sex). The typical-value ratios below are for a 79 kg patient.

```{r fig4}
ss_total_auc <- function(covs) {
  s <- rxode2::rxSolve(mod_typ, make_events(1L, seq(0, 139 * 24, by = 24), 6,
                                            seq(139 * 24, 140 * 24, by = 0.1), covs),
                       returnType = "data.frame") |> to_nM()
  trap(s$time, s$Total)
}
ref <- ss_total_auc(male_white(79))
groups <- list(
  Black = data.frame(WT = 79, SEXF = 0, RACE_BLACK = 1, RACE_ASIAN = 0, RACE_JAPANESE = 0),
  Asian = data.frame(WT = 79, SEXF = 0, RACE_BLACK = 0, RACE_ASIAN = 1, RACE_JAPANESE = 0),
  Japanese = data.frame(WT = 79, SEXF = 0, RACE_BLACK = 0, RACE_ASIAN = 0, RACE_JAPANESE = 1),
  Female = data.frame(WT = 79, SEXF = 1, RACE_BLACK = 0, RACE_ASIAN = 0, RACE_JAPANESE = 0)
)
fig4 <- data.frame(group = names(groups),
                   ratio_total_auc = vapply(groups, ss_total_auc, numeric(1)) / ref)
knitr::kable(fig4, digits = 3, caption = "Typical-value ratio of steady-state Total CAR AUC0-24 to a 79 kg White male.")
stopifnot(all(abs(fig4$ratio_total_auc - 1) < 0.36))
```

The directions match Figure 4 of the paper: lower exposure in Black patients,
higher in Asian and Japanese patients, and higher in women.

# Population simulation and NCA (Table 3)

Table 3 summarises individual steady-state exposure at 6 mg/day, computed from
each patient's post hoc parameters. A virtual cohort of 200 patients is drawn
from the Supplemental Table 2 demographics and dosed for 12 weeks.

```{r cohort}
set.seed(20211103)
n_sub <- 200
draw_wt <- function(n) {
  # Reject-and-redraw inside the observed range rather than clamping.
  w <- rnorm(n, 78.9, 18.7)
  bad <- w < 33.1 | w > 155.1
  while (any(bad)) {
    w[bad] <- rnorm(sum(bad), 78.9, 18.7)
    bad <- w < 33.1 | w > 155.1
  }
  w
}
race <- sample(c("White", "Black", "Asian", "Japanese", "Other"), n_sub, replace = TRUE,
               prob = c(45.6, 34.9, 14.3, 1.7, 3.5))
cohort <- data.frame(
  id = seq_len(n_sub), WT = draw_wt(n_sub), SEXF = rbinom(n_sub, 1, 0.336),
  RACE_BLACK = as.integer(race == "Black"), RACE_ASIAN = as.integer(race == "Asian"),
  RACE_JAPANESE = as.integer(race == "Japanese")
)
events <- do.call(rbind, lapply(seq_len(n_sub), function(i) {
  make_events(i, seq(0, 83 * 24, by = 24), 6,
              c(0, seq(83 * 24, 84 * 24, by = 0.5)), cohort[i, -1])
})) |>
  dplyr::mutate(treatment = "6 mg QD")
stopifnot(!anyDuplicated(events[, c("id", "time", "evid")]))
```

```{r simulate}
sim <- rxode2::rxSolve(mod, events = events, keep = "treatment", returnType = "data.frame")
sim_nM <- sim |>
  dplyr::transmute(
    id, time, treatment,
    CAR = Cc / mw[["CAR"]], DCAR = Cc_dcar / mw[["DCAR"]], DDCAR = Cc_ddcar / mw[["DDCAR"]]
  ) |>
  tidyr::pivot_longer(c(CAR, DCAR, DDCAR), names_to = "analyte", values_to = "Cc")

sim_nM |>
  dplyr::filter(time >= 83 * 24) |>
  dplyr::group_by(analyte, time) |>
  dplyr::summarise(Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
                   .groups = "drop") |>
  ggplot(aes(time - 83 * 24, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~analyte, scales = "free_y") +
  labs(x = "Time after the 84th daily 6 mg dose (h)", y = "Concentration (nM)",
       title = "Simulated steady-state profiles (median, 5th-95th percentile)")
```

```{r pknca}
sim_nca <- sim_nM |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment, analyte)
conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + analyte + id)
dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)
intervals <- data.frame(start = 83 * 24, end = 84 * 24,
                        cmax = TRUE, cmin = TRUE, auclast = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

## Comparison against published steady-state exposure

`ncaComparisonTable()` compares the simulated median with the Table 3 median.

```{r nca_compare}
published <- tibble::tribble(
  ~treatment, ~analyte, ~cmax, ~cmin, ~auclast,
  "6 mg QD", "CAR", 35.4, 19.9, 641.0,
  "6 mg QD", "DCAR", 8.1, 6.1, 171.1,
  "6 mg QD", "DDCAR", 55.3, 55.3, 1325.9
)
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = c("treatment", "analyte"),
  units = c(cmax = "nM", cmin = "nM", auclast = "nM*h"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Simulated vs Table 3 median steady-state exposure at 6 mg/day. * differs from reference by >20%.")
```

```{r nca_gate}
sim_med <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "auclast")) |>
  dplyr::group_by(analyte, PPTESTCD) |>
  dplyr::summarise(sim = median(PPORRES), .groups = "drop") |>
  dplyr::left_join(
    published |> tidyr::pivot_longer(c(cmax, auclast), names_to = "PPTESTCD", values_to = "ref") |>
      dplyr::select(analyte, PPTESTCD, ref),
    by = c("analyte", "PPTESTCD")
  ) |>
  dplyr::mutate(pct_diff = 100 * (sim - ref) / ref)
# Centre-of-distribution gate on a random cohort. With a 1500-subject cohort
# the model medians sat 8-15% above Table 3 for all three analytes (Table 3
# is built from post hoc estimates of 2599 patients, not from fresh draws);
# 30% absorbs the cohort-to-cohort spread of a 200-subject median while still
# failing on a mis-transcribed clearance or a dose/unit error.
stopifnot(nrow(sim_med) == 6, all(abs(sim_med$pct_diff) < 30))
```

Table 3 lists identical Cmax,ss and Cmin,ss statistics for DDCAR (mean 65.6,
median 55.3 nM in both rows), which is not possible for a drug with any
within-interval fluctuation; the DDCAR Cmin,ss reference above is therefore the
paper's Cmax,ss value repeated, and that row should be read with this in mind.
DDCAR does fluctuate little over a dosing interval, so the two numbers are
close in the simulation too.

# Initial-dataset model

The initial-dataset model has a two-compartment cariprazine and one-compartment
DCAR and DDCAR, with ideal body weight (reference 64.5 kg), age, race and sex
covariates entering mostly as additive shifts. It was fitted only to samples
within 25 h of a dose, so its terminal phases are not supported beyond that
window. The panel below compares the two models for a typical 40-year-old White
man (79 kg, ideal body weight 64.5 kg) after the first dose and on day 84.

```{r initial, fig.height = 4.5}
mod_init_typ <- rxode2::zeroRe(rxode2::rxode2(readModelDb("Periclou_2021_cariprazine_initial")))
cov_both <- data.frame(WT = 79, IBW = 64.5, AGE = 40, SEXF = 0, RACE_BLACK = 0,
                       RACE_ASIAN = 0, RACE_JAPANESE = 0)
obs_cmp <- c(seq(0, 24, by = 0.25), seq(83 * 24, 84 * 24, by = 0.25))
ev_cmp <- make_events(1L, seq(0, 83 * 24, by = 24), 6, obs_cmp, cov_both)
cmp_models <- dplyr::bind_rows(
  rxode2::rxSolve(mod_typ, ev_cmp, returnType = "data.frame") |> to_nM() |>
    dplyr::mutate(model = "Updated (final)"),
  rxode2::rxSolve(mod_init_typ, ev_cmp, returnType = "data.frame") |> to_nM() |>
    dplyr::mutate(model = "Initial dataset")
) |>
  dplyr::mutate(period = ifelse(time < 24.5, "Day 1", "Day 84"),
                tad = ifelse(time < 24.5, time, time - 83 * 24)) |>
  tidyr::pivot_longer(c(CAR, DCAR, DDCAR), names_to = "analyte", values_to = "conc")
ggplot(cmp_models, aes(tad, conc, colour = model)) +
  geom_line() +
  facet_grid(analyte ~ period, scales = "free_y") +
  labs(x = "Time after dose (h)", y = "Concentration (nM)", colour = NULL,
       title = "Initial-dataset vs updated model, 6 mg once daily")
```

```{r initial-mass-balance}
ev_mbi <- make_events(1L, seq(0, 139 * 24, by = 24), 6,
                      seq(139 * 24, 140 * 24, by = 0.05), cov_both)
mbi <- rxode2::rxSolve(mod_init_typ, ev_mbi, returnType = "data.frame")
mbi_tab <- data.frame(
  analyte = c("CAR", "DCAR", "DDCAR"),
  auc_tau = c(trap(mbi$time, mbi$Cc), trap(mbi$time, mbi$Cc_dcar), trap(mbi$time, mbi$Cc_ddcar)),
  dose_over_cl = 6000 / c(22.8, 70.9, 6.74)
) |>
  dplyr::mutate(pct_diff = 100 * (auc_tau - dose_over_cl) / dose_over_cl)
knitr::kable(mbi_tab, digits = 2,
             caption = "Initial-dataset model: steady-state AUC over 24 h vs Dose/CL (ng*h/mL) at the reference covariates.")
stopifnot(max(abs(mbi_tab$pct_diff)) < 0.5)
```

# Assumptions and deviations

- **Residual error is not reported.** Table 2 footnote f says residual
  variability was estimated separately for the phase 1 and phase 2/3 studies,
  but neither the error model nor its values appear in the paper or supplement.
  Both packaged models carry proportional errors fixed to zero, so they simulate
  typical and between-subject variability only and are not suitable for
  re-estimation as distributed.
- **IIV variances.** Table 2 and Supplemental Table 3 report IIV as %CV; they
  were converted with `omega^2 = log(1 + CV^2)`, and no correlations are
  reported, so the random effects are independent. IIV on `Ka`, cariprazine
  `Vc/F` and DCAR `Vc/F` was estimated from phase 1 patients only (Table 2
  footnote b; shrinkage 76.5% on `Ka`); the packaged model applies it to every
  simulated subject.
- **Mass chain, no molecular-weight term.** The printed equations transfer
  cariprazine elimination to DCAR, and DCAR elimination to DDCAR, amount for
  amount, and all DCAR and DDCAR parameters are apparent (Section 2.4.1). The
  packaged model follows the equations literally, in mg of dose with
  concentrations in ng/mL. Had the authors instead worked in molar amounts,
  predicted DCAR and DDCAR concentrations would be 3.3% and 6.6% lower. The
  paper's own summary numbers do not settle this: the Figure 5c DDCAR plateau
  is closer to the molar reading, while the DDCAR-to-cariprazine ratios of the
  Table 3 means and medians are closer to the mass reading. Either reading is
  well inside the between-subject variability.
- **First-dose shift.** `FD` is built inside the model from `dosenum()` rather
  than read from a data column. In a NONMEM dataset the flag switches on
  the record after which it is set; here it switches at the second dose itself.
- **Figure 5 typical weight.** The text describes the Figure 5 patient as
  79 kg, but the first-dose profile reproduces to within 3% only at 84 kg, the
  typical male of Figures 6 and 7, so Figure 5 is replicated at 84 kg.
- **Figure 5a at 24 h is not reproduced.** At 24 h after the first dose the
  model gives 5.5 nM cariprazine, 1.4 nM DCAR and 0.3 nM DDCAR, against 2.2, 1.1
  and 0.5 nM in the figure, while every earlier point agrees closely. With
  Q4/F = 100 L/h and VP2/F = 501 L the second peripheral compartment
  re-equilibrates within hours and the first-dose Vc/F is 1021 L, so no
  reading of the printed parameters produces the figure's 3-fold fall between
  12 and 24 h. Placing the first-dose shift on VP2/F instead of VP1/F, or
  switching it off over the 12-24 h interval as a NONMEM record at 24 h would,
  does not reproduce it either. The deviation is recorded and excluded from
  the gate.
- **Figure 5 metabolite curves.** Figure 5c's Total curve sits about 130 nM*h
  below the sum of its own three analyte curves, and its DCAR plateau
  (232 nM*h) cannot be reconciled with the DCAR concentrations of Figure 5b
  (about 5.5-7.8 nM, i.e. 130-190 nM*h per day). At steady state the model
  gives DCAR 181 nM*h and DDCAR 1576 nM*h for the 84 kg patient, against
  232 and 1430 in the figure. Only the cariprazine curve, which is consistent
  across all three panels, is gated on Figure 5.
- **Table 3 DDCAR Cmin,ss** repeats the Cmax,ss statistics exactly and is
  taken as printed.
- **Race indicators.** In the updated analysis race had five mutually
  exclusive levels. `RACE_ASIAN` is the non-Japanese Asian group (mainly
  patients at Indian sites) and must be 0 for Japanese patients, who carry
  `RACE_JAPANESE = 1`. White and Other form the reference.
- **Initial-dataset Equation A6.** Supplemental Equation A6 prints the DDCAR
  central volume as
  `2220 + 27.4 x (Age - 40) + 39.7 x (IBW - 64.5) x (1180 x Black)`, which
  would remove the ideal-body-weight term entirely for non-Black patients.
  Supplemental Table 3 lists 1180 L as an "Additional shift in black patients
  (L)", so the Black term is encoded as an additive `+ 1180 x Black`.
- **Initial-dataset IIV and IBW.** The between-subject variances of the
  initial-dataset model are applied multiplicatively (log-normal) to the
  additive covariate expressions. The ideal-body-weight formula is not given in
  the paper.
- **Virtual cohort.** Body weight was drawn from a normal distribution with the
  Supplemental Table 2 mean and SD and redrawn inside the observed range; race
  and sex follow the Supplemental Table 2 proportions. No correlation between
  weight, sex and race was imposed.
- **Errata.** No correction or erratum is linked to this article
  (PMID 33141308) in Europe PMC as of 2026-09-27.
