---
title: "TV-46000 long-acting subcutaneous risperidone (Perlstein 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TV-46000 long-acting subcutaneous risperidone (Perlstein 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r citation, echo = FALSE, results = "asis"}
m <- rxode2::rxode(readModelDb("Perlstein_2025_risperidone_tv46000"))
cat("* Citation: ", m$reference, "\n\n", sep = "")
cat("* Description: ", m$description, "\n\n", sep = "")
```

* Article: <https://doi.org/10.1007/s40120-025-00723-z>
* Supplement (Table S1, Figures S1-S3): <https://doi.org/10.1007/s40120-025-00723-z>

TV-46000 is a long-acting subcutaneous antipsychotic (LASCA): risperidone
suspended in a copolymer-based delivery vehicle at roughly 360 mg/mL and injected
subcutaneously into the abdomen or the back of the upper arm once monthly (q1m)
or once every 2 months (q2m). Perlstein 2025 pooled three phase 1 studies with
the two phase 3 trials RISE and SHINE and fitted a **sequential
parent-metabolite** population PK model: risperidone first, then
9-hydroxyrisperidone (paliperidone) driven by the parent's individual post-hoc
predictions.

The clinically reported quantity is the **total active moiety** (TAM), because
risperidone and 9-hydroxyrisperidone are equipotent at the dopamine D2 receptor.
Perlstein 2025 defines it in the Introduction as the "sum of concentrations of
risperidone and its active metabolite [9-OH risperidone] corrected by molecular
weight", i.e. in risperidone-equivalent units.

Three risperidone models were already in the library when this one was added, and
the contrast is worth stating because it decides which one a user should reach
for:

* `modellib("Ivaturi_2017_RBP_7000")` -- the **structural ancestor** of this
  model. Perlstein 2025 Methods: "The starting point of the current analysis was
  the model reported by Ivaturi et al., which described sc absorption by one
  depot dose and a double first-order absorption route." Ivaturi 2017 is a
  different LAI product (RBP-7000 / Atrigel) and pairs its PK with a PANSS
  exposure-response model.
* `modellib("Wang_2024_risperidone_consta")` and
  `modellib("Wang_2024_risperidone_rykindo")` -- intramuscular LAI formulations
  modelled on the active moiety directly, without splitting parent from
  metabolite.
* `modellib("Feng_2008_risperidone")`, `modellib("Sherwin_2012_risperidone")` --
  oral risperidone.

# Population

The analysis dataset comprised **692 participants** with at least one measurable
plasma concentration after a TV-46000 injection: 267 from three phase 1 studies
(50 healthy volunteers in RISPE1ZG15EU, 97 in SAD-10055, 120 in BA-10148) and 425
from the two phase 3 trials (352 in RISE, NCT03503318; 73 in SHINE,
NCT03893825). Two further participants were excluded for self-administration of
risperidone and 41 because every PK sample was below the limit of quantification
(Perlstein 2025 Results "Dataset" and Table S1).

Baseline demographics (Perlstein 2025 Table 2, Total Overall column, n = 733
participants with baseline data) were: mean age 47.4 years (SD 10.9, range
16-65), 70.4% male, 64.7% Black or African American / 31.9% White / 1.4% Asian,
mean weight 86.3 kg (SD 16.3, range 42-132), mean BMI 28.7 kg/m^2 (SD 4.8, range
18-38) and mean creatinine clearance 121 mL/min (SD 34.3). Across 3287
injections, 68.5% were abdominal and 31.5% into the upper arm, and 77.8% were
delivered from a vial versus 22.2% from a prefilled syringe. The mean injection
volume was 0.303 mL (SD 0.144).

Patients over 65 years were excluded from the source trials, and the Discussion
flags the male and Black/African American skew as the analysis's main
generalizability limitation.

The same information is available programmatically:

```{r population, echo = TRUE}
str(readModelDb("Perlstein_2025_risperidone_tv46000")()$population, max.level = 1)
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Perlstein_2025_risperidone_tv46000.R` carries an
in-file comment naming its source location. They are collected here for review.

| Equation / parameter | Value | Source location |
|---|---|---|
| Model structure (depot with parallel `ka1` / `ka2` exits, 5-compartment `ktr` transit chain, parent -> metabolite) | n/a | Figure 1 (model schematic); Results, "Risperidone (Parent) PK Model" |
| Continuous covariate form `Pi = P * (COVi / COVmedian)^theta_i` | n/a | Methods, "Population PK Model Development" |
| Categorical covariate form `Pi = P * (1 + theta_i)^COVi` | n/a | Methods, "Population PK Model Development" |
| `lcl` (CL/F) | 14.3 L/h | Table 3, risperidone (parent) |
| `lvc` (V/F) | 66.3 L | Table 3, risperidone (parent) |
| `lka1` (KA1) | 0.000632 1/h | Table 3, risperidone (parent) |
| `lka2` (KA2) | 0.000408 1/h | Table 3, risperidone (parent) |
| `lktr` (KTR) | 0.0252 1/h | Table 3, risperidone (parent) |
| `e_bmi_ka1` (KA1BMI1) | -1.1 | Table 3 |
| `e_dose_ka1` (KA1INJV1) | -0.384 | Table 3 |
| `e_injsite_arm_ka1` (KA1ADMSITE1) | 0.331 | Table 3; Results ("33% higher KA1" in the upper arm) |
| `e_bmi_ka2` (KA2BMI1) | 1.7 | Table 3 |
| `lcl_9oh` (CLMO) | 5.78 L/h | Table 3, 9-OH risperidone (metabolite) |
| `lvc_9oh` (VMO) | 95.7 L | Table 3, 9-OH risperidone (metabolite) |
| FRMET fixed to 1 | 1 | Table 3 footnote |
| `etalka2` / `etalka1` block | 254.2% / 51.0% CV, correlation 42.8% | Table 3, random effects |
| `etalcl` | 82.3% CV | Table 3, random effects |
| `etalcl_9oh` | 65.1% CV | Table 3, random effects |
| `propSd` (parent EP) | 40.5% | Table 3, residual error; Methods (log-transformed both sides) |
| `propSd_9oh` (metabolite EP) | 38.3% | Table 3, residual error |
| `lemax` (D2RO Emax) | 100% | Methods, "Prediction of Individual Exposure Parameters..." |
| `kd` (D2RO) | 10.1 ng/mL | Methods, "Prediction of Individual Exposure Parameters..." |
| TAM = risperidone + molecular-weight-corrected 9-OH-risperidone | 410/426 | Introduction (definition of TAM); the two molecular weights are not printed -- see Assumptions |
| Injection volumes 0.035 / 0.07 / 0.139 mL at 12.5 / 25 / 50 mg | n/a | Results, "Risperidone (Parent) PK Model"; 0.278 mL at 100 mg from Table 2 (BA-10148) |
| 28-day month for simulations | n/a | Methods, "Model-Based PK Simulations ... for Different Dosing Regimens" |

## Reading the IIV percentages

Table 3 reports each interindividual variability as a bare percentage. Two
conventions are in circulation -- `%CV = 100 * sqrt(exp(omega^2) - 1)` and the
approximation `%CV = 100 * omega` -- and they disagree badly for the very large
KA2 term (`omega^2` = 2.010 versus 6.462). The table's own confidence intervals
settle it. Propagating the printed `%RSE` as a relative standard error **on
`omega`** through the first (exact-lognormal) transform reproduces the printed
95% CIs closely, while the second does not:

```{r iiv-convention, echo = FALSE}
iiv <- tibble::tribble(
  ~Parameter, ~cv,   ~rse,  ~lo_pub, ~hi_pub,
  "KA2",      254.2, 0.052, 198.8,   320.1,
  "KA1",       51.0, 0.060,  44.0,    57.5,
  "CL",        82.3, 0.033,  75.3,    89.3,
  "CLMO",      65.1, 0.040,  58.9,    71.1
) |>
  mutate(
    om2      = log(1 + (cv / 100)^2),
    om       = sqrt(om2),
    se_om2   = 2 * om * (rse * om),
    dlncv    = 0.5 * exp(om2) / (exp(om2) - 1),
    se_lncv  = dlncv * se_om2,
    A_lo     = cv * exp(-1.96 * se_lncv), A_hi = cv * exp(1.96 * se_lncv),
    B_lo     = cv * exp(-1.96 * rse),     B_hi = cv * exp(1.96 * rse)
  )
iiv |>
  transmute(
    Parameter,
    `Published %CV (95% CI)` = sprintf("%.1f (%.1f-%.1f)", cv, lo_pub, hi_pub),
    `omega^2 if lognormal`   = sprintf("%.3f", om2),
    `CI implied, lognormal`  = sprintf("%.1f-%.1f", A_lo, A_hi),
    `CI implied, omega=CV`   = sprintf("%.1f-%.1f", B_lo, B_hi)
  ) |>
  knitr::kable()
```

The lognormal reading is used, so `omega^2 = log(1 + CV^2)` throughout. It is
also the convention the upstream `Ivaturi_2017_RBP_7000` model uses.

# Structural check: steady-state mass balance

Before any population simulation, a typical-value (`zeroRe`) solve confirms that
the packaged ODE system conserves mass the way the published parameterisation
requires. Because the depot's only exits are the two absorption routes and
FRMET is fixed to 1, every milligram injected must eventually appear as parent
elimination and then as metabolite elimination. At steady state that forces two
exact identities over one dosing interval:

$$AUC_{ss,\text{risperidone}} = \frac{1000 \cdot D}{CL/F}
\qquad
AUC_{ss,\text{9-OH}} = \frac{1000 \cdot D}{CLMO}$$

```{r mass-balance}
mod  <- readModelDb("Perlstein_2025_risperidone_tv46000")
tmod <- rxode2::zeroRe(mod)

TAU_Q1M <- 28 * 24   # h; the paper assumes a 28-day month
TAU_Q2M <- 56 * 24   # h
MG_PER_ML <- 360     # TV-46000 suspension strength (see model file notes)

typicalProfile <- function(dose_mg, tau, n_dose, bmi = 28.7, arm = 0L,
                           step = 1, window_only = TRUE) {
  grid <- if (window_only) {
    seq((n_dose - 1) * tau, n_dose * tau, by = step)
  } else {
    seq(0, n_dose * tau, by = step)
  }
  ev <- bind_rows(
    data.frame(time = (seq_len(n_dose) - 1) * tau, amt = dose_mg,
               cmt = "depot", evid = 1L, dvid = NA_integer_),
    data.frame(time = grid, amt = NA_real_, cmt = NA_character_,
               evid = 0L, dvid = 1L)
  ) |>
    mutate(BMI = bmi, INJSITE_ARM = arm,
           DOSE_TV46000_ML = dose_mg / MG_PER_ML) |>
    arrange(time, desc(evid))
  rxode2::rxSolve(tmod, ev, omega = NA, useLinCmt = FALSE,
                  addDosing = FALSE, returnType = "data.frame")
}

trapz <- function(x, y) sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)

ss75 <- typicalProfile(75, TAU_Q1M, n_dose = 20)
auc_parent <- trapz(ss75$time, ss75$Cc)
auc_met    <- trapz(ss75$time, ss75$Cc_9oh)

massBalance <- data.frame(
  Analyte    = c("Risperidone", "9-OH-risperidone"),
  Simulated  = c(auc_parent, auc_met),
  Identity   = c(1000 * 75 / 14.3, 1000 * 75 / 5.78)
) |>
  mutate(`% diff` = 100 * (Simulated / Identity - 1))
knitr::kable(massBalance, digits = c(0, 1, 1, 4))

stopifnot(all(abs(massBalance$`% diff`) < 0.1))
```

Both identities hold to better than 0.1%, which pins the depot split, the
transit chain and the FRMET = 1 metabolite coupling simultaneously: any error in
the release structure would leak or duplicate mass and break them.

# Virtual cohort

Original subject-level data are not public. Perlstein 2025 bootstrapped
"demographic covariates ... from a random sample of 500 participants from the
TV-46000 phase 3 population" for its simulations; the cohort below approximates
that phase 3 population from Table 2. BMI is drawn from the phase 3 mean and SD
and truncated to the published range, and the injection site is assigned per
subject at the RISE arm frequency.

`DOSE_TV46000_ML` is derived from the milligram dose using the suspension
strength implied by the paper's own printed volumes (12.5 mg / 0.035 mL,
25 mg / 0.07 mL, 50 mg / 0.139 mL, 100 mg / 0.278 mL -- all ~360 mg/mL).

```{r cohort}
NSUB <- 100L
set.seed(20250323)

cohort <- data.frame(
  id          = seq_len(NSUB),
  BMI         = pmin(pmax(rnorm(NSUB, mean = 29.0, sd = 4.9), 18), 38),
  INJSITE_ARM = rbinom(NSUB, size = 1L, prob = 0.344)
)

summary(cohort$BMI)
table(Arm = cohort$INJSITE_ARM)
```

```{r sim-helper}
buildEvents <- function(dose_mg, tau, n_dose, grid, cov = cohort) {
  bind_rows(
    tidyr::crossing(id = cov$id, time = (seq_len(n_dose) - 1) * tau) |>
      mutate(amt = dose_mg, cmt = "depot", evid = 1L, dvid = NA_integer_),
    tidyr::crossing(id = cov$id, time = grid) |>
      mutate(amt = NA_real_, cmt = NA_character_, evid = 0L, dvid = 1L)
  ) |>
    left_join(cov, by = "id") |>
    mutate(DOSE_TV46000_ML = dose_mg / MG_PER_ML) |>
    arrange(id, time, desc(evid))
}

SIM_SEED <- 20250323L

simRegimen <- function(dose_mg, tau, n_dose, grid, cov = cohort) {
  # Common random numbers: reseeding before every solve makes each regimen and
  # each injection-site scenario draw the SAME etas, so a between-scenario
  # difference reflects the scenario rather than the sampling. Without this,
  # the injection-site comparison below is two independent eta draws and its
  # difference is pure Monte Carlo noise.
  set.seed(SIM_SEED)
  rxode2::rxSolve(mod, buildEvents(dose_mg, tau, n_dose, grid, cov),
                  useLinCmt = FALSE, addDosing = FALSE,
                  returnType = "data.frame")
}
```

Note the two rxode2 requirements this model imposes on every event table. It
declares **two** endpoints (`Cc` and `Cc_9oh`), so observation rows must identify
which endpoint they belong to -- here by `dvid = 1L` with `cmt = NA_character_`,
while dose rows carry the real ODE state `cmt = "depot"`. And `useLinCmt = FALSE`
is passed to every solve, because the automatic ODE-to-`linCmt()` conversion
corrupts the endpoint mapping.

# Simulated steady-state TAM profiles

Replicates Figure 2 of Perlstein 2025 (simulated TAM concentration-time profiles
for q1m and q2m dosing), shown here over the steady-state dosing interval.

```{r sim-main}
regimens <- bind_rows(
  data.frame(regimen = "q1m", dose = c(50, 75, 100, 125),
             tau = TAU_Q1M, n_dose = 13L),
  data.frame(regimen = "q2m", dose = c(100, 150, 200, 250),
             tau = TAU_Q2M, n_dose =  7L)
) |>
  mutate(treatment = paste0("TV-46000 ", dose, " mg ", regimen))

sims <- lapply(seq_len(nrow(regimens)), function(i) {
  r <- regimens[i, ]
  grid <- sort(unique(c(seq(0, r$tau, by = 6),
                        seq((r$n_dose - 1) * r$tau, r$n_dose * r$tau, by = 6))))
  simRegimen(r$dose, r$tau, r$n_dose, grid) |>
    mutate(regimen = r$regimen, dose = r$dose, treatment = r$treatment,
           tau = r$tau, n_dose = r$n_dose)
}) |> bind_rows()

ssBand <- sims |>
  filter(time >= (n_dose - 1) * tau) |>
  mutate(tad = time - (n_dose - 1) * tau, days = tad / 24) |>
  group_by(regimen, treatment, dose, days) |>
  summarise(p5 = quantile(TAM, 0.05), p50 = median(TAM),
            p95 = quantile(TAM, 0.95), .groups = "drop")

ggplot(ssBand, aes(days, p50, colour = factor(dose), fill = factor(dose))) +
  geom_ribbon(aes(ymin = p5, ymax = p95), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~regimen, scales = "free_x") +
  labs(x = "Days since the steady-state dose", y = "TAM (ng/mL)",
       colour = "Dose (mg)", fill = "Dose (mg)",
       title = "Steady-state total active moiety",
       subtitle = "Median with 5th-95th percentile band; replicates Figure 2 of Perlstein 2025") +
  theme_bw()
```

# Therapeutic concentrations within 24 hours of the first dose

Perlstein 2025 Abstract and Key Summary Points: "TV-46000 reaches therapeutic
plasma TAM concentrations (>= 10 ng/mL) within 24 h following first dose
administration". At 24 h the transit chain has delivered essentially nothing, so
this claim is carried entirely by the **fast direct route** -- it is a sharp,
almost independent check on `ka1` and its covariate reference values.

```{r first-dose}
# At 24 h after the first injection the dosing interval is irrelevant -- only the
# dose strength (through the injection-volume covariate on ka1) matters -- so the
# comparison is made over the distinct strengths.
typical24 <- lapply(sort(unique(regimens$dose)), function(d) {
  s <- typicalProfile(d, TAU_Q1M, n_dose = 1, step = 1, window_only = FALSE)
  data.frame(dose = d, typical = s$TAM[which.min(abs(s$time - 24))])
}) |> bind_rows()

firstDose <- sims |>
  filter(time == 24) |>
  group_by(dose) |>
  summarise(popMedian = median(TAM), fracOver = mean(TAM >= 10),
            .groups = "drop") |>
  left_join(typical24, by = "dose")

firstDose |>
  transmute(`Dose (mg)` = dose,
            `Typical-participant TAM (ng/mL)` = typical,
            `Population median TAM (ng/mL)` = popMedian,
            `Fraction of cohort >= 10 ng/mL` = fracOver) |>
  knitr::kable(digits = 2)
```

The model reproduces the claim at the mid-to-upper strengths and falls modestly
short at the lowest ones: the typical participant clears 10 ng/mL from 100 mg
upward but reaches only about 7.6 and 9.8 ng/mL at 50 and 75 mg, where roughly
half the simulated cohort is above the threshold. The paper states the claim for
the programme without a dose qualifier, so this is a genuine dose-resolved
refinement of it rather than agreement. Two mitigating points, neither of which
the packaged model carries: phase 3 patients were stabilized on oral risperidone
for 12 weeks before the first injection (Methods notes the residual contribution
is "marginal" by 24 h), and the published claim is a population statement, which
the median column satisfies from 75 mg upward.

The population median runs above the typical participant because interindividual
variability on `ka1` is lognormal and the 24-h concentration is close to linear
in `ka1`, so the covariate-free typical value is not the median of the cohort.

# PKNCA validation and comparison with Table 4

Perlstein 2025 Table 4 reports model-derived **median** steady-state TAM
exposures for all eight approved dosing regimens. That table is a whole-model
answer key: 32 numbers spanning both dosing intervals and the full dose range,
each of which depends jointly on the release structure, the covariate model and
the metabolite coupling.

NCA is computed with PKNCA over the steady-state dosing interval. The
concentration column is named `Cc` to match the package convention, but it holds
the **total active moiety**, which is the analyte Table 4 reports.

```{r pknca}
ncaConc <- sims |>
  filter(time >= (n_dose - 1) * tau) |>
  transmute(id, treatment, time, Cc = TAM) |>
  filter(!is.na(Cc))

ncaDose <- regimens |>
  tidyr::crossing(id = cohort$id) |>
  transmute(id, treatment, time = (n_dose - 1) * tau, amt = dose)

concObj <- PKNCA::PKNCAconc(ncaConc, Cc ~ time | treatment + id,
                            concu = "ng/mL", timeu = "h")
doseObj <- PKNCA::PKNCAdose(ncaDose, amt ~ time | treatment + id, doseu = "mg")

intervals <- regimens |>
  tidyr::crossing(id = cohort$id) |>
  transmute(treatment, id,
            start = (n_dose - 1) * tau, end = n_dose * tau,
            cmax = TRUE, tmax = TRUE, cmin = TRUE,
            auclast = TRUE, cav = TRUE) |>
  as.data.frame()

ncaRes <- PKNCA::pk.nca(PKNCA::PKNCAdata(concObj, doseObj, intervals = intervals))
```

`ctrough` is not used here: PKNCA can return it as all-`NA` for a steady-state
interval, and the paper's `C_trough,ss` is by definition the concentration at the
end of the interval, which is read straight off the solved profile instead.

```{r nca-table}
ctroughSim <- sims |>
  filter(time == n_dose * tau) |>
  transmute(id, treatment, PPTESTCD = "ctrough", PPORRES = TAM)

simNca <- bind_rows(
  as.data.frame(ncaRes$result) |> select(treatment, id, PPTESTCD, PPORRES),
  ctroughSim
)

published <- regimens |>
  transmute(treatment,
            auclast = c(13306, 19951, 26521, 33141, 26829, 40152, 53321, 66634),
            cav     = c( 19.8,  29.7,  39.5,  49.3,  20.0,  29.9,  39.7,  49.6),
            cmax    = c( 25.1,  36.4,  47.5,  58.7,  32.6,  47.2,  61.4,  76.0),
            ctrough = c( 12.1,  19.5,  26.8,  34.1,   7.7,  12.8,  17.6,  22.8)) |>
  as.data.frame()

cmpTable <- nlmixr2lib::ncaComparisonTable(
  simNca, published, by = "treatment",
  params = c("auclast", "cmax", "cav", "ctrough"),
  units  = c(auclast = "ng*h/mL", cmax = "ng/mL",
             cav = "ng/mL", ctrough = "ng/mL")
)
knitr::kable(cmpTable, digits = 1)
```

```{r nca-footnote, echo = FALSE, results = "asis"}
fn <- attr(cmpTable, "footnote")
if (!is.null(fn) && nzchar(fn)) cat("\n", fn, "\n", sep = "")
pctDiff <- abs(as.numeric(gsub("[^0-9.-]", "", cmpTable[["% diff"]])))
cat("\nLargest absolute difference across the ", nrow(cmpTable),
    " comparisons: ", sprintf("%.1f%%", max(pctDiff)),
    "; median absolute difference ", sprintf("%.1f%%", median(pctDiff)), ".\n",
    sep = "")
```

Every one of the `r nrow(cmpTable)` comparisons falls inside the 20% flagging
tolerance, and the agreement is far tighter than that: the exposure metrics
`AUC_ss` and `C_avg,ss` reproduce within about 1% on all eight regimens, and
`C_max,ss` within about 3%.

The gradient across the four parameters is itself informative, because it tracks
how much of each one is fixed by mass balance rather than by the release model.
`AUC_ss` and `C_avg,ss` are pinned by `1000 * D / CL` and `1000 * D / CLMO`
(the identity checked above), so they test the clearance estimates and little
else -- agreeing to 1% is expected once those transcribe correctly. `C_max,ss`
and especially `C_trough,ss` depend on the *shape* of the release profile, which
is where the two covariate centring values this paper never published
(see Assumptions) and the exact BMI distribution of the virtual cohort actually
bite. That the shape-sensitive metrics still land within a few percent is the
part of this comparison that carries evidence about the release structure.

An earlier draft of this vignette drew independent random effects for each
regimen and each injection-site scenario; that alone produced deviations up to
14% and an apparent 3% injection-site effect, both of which were sampling noise.
The common-random-numbers seeding in `simRegimen()` above removes it.

# Dose proportionality and dosing-interval interchangeability

Perlstein 2025 Results: "At corresponding doses of TV-46000, TAM AUC_ss following
q2m dosing was twice that of the TAM AUC_ss following q1m dosing, as TV-46000
showed linearity in PK. Therefore, the average daily exposures with q1m and q2m
were similar, supporting dosing interval interchangeability." The corresponding
pairs are the q2m strength that is double the q1m strength.

```{r proportionality}
ssExposure <- as.data.frame(ncaRes$result) |>
  filter(PPTESTCD == "auclast") |>
  left_join(regimens, by = "treatment") |>
  group_by(regimen, dose, tau) |>
  summarise(auc = median(PPORRES), .groups = "drop") |>
  mutate(daily = auc / (tau / 24))

pairedDoses <- data.frame(q1m = c(50, 75, 100, 125), q2m = c(100, 150, 200, 250)) |>
  left_join(ssExposure |> filter(regimen == "q1m") |>
              select(q1m = dose, auc1 = auc, daily1 = daily), by = "q1m") |>
  left_join(ssExposure |> filter(regimen == "q2m") |>
              select(q2m = dose, auc2 = auc, daily2 = daily), by = "q2m") |>
  mutate(`AUCss ratio (q2m / q1m)` = auc2 / auc1,
         `Daily AUCss ratio` = daily2 / daily1)

pairedDoses |>
  transmute(`q1m dose (mg)` = q1m, `q2m dose (mg)` = q2m,
            `AUCss ratio (q2m / q1m)`, `Daily AUCss ratio`) |>
  knitr::kable(digits = 2)
```

The AUC_ss ratio is close to 2 at every matched pair (median
`r sprintf("%.2f", median(pairedDoses[["AUCss ratio (q2m / q1m)"]]))`) while the daily
AUC_ss ratio is close to 1 (median
`r sprintf("%.2f", median(pairedDoses[["Daily AUCss ratio"]]))`), reproducing the
paper's interchangeability claim.

# Time to steady state

Perlstein 2025 Results and Conclusion: steady state is "approached by
approximately 3 months after initiation (86% and 88% of steady-state TAM exposure
for q1m and q2m, respectively) and fully attained by approximately 6 months
(i.e., > 90% of steady-state TAM exposure)". Each participant's per-interval AUC
is compared with their own final-interval AUC.

```{r steady-state}
attainment <- lapply(list(list("q1m", 75, TAU_Q1M, 13L),
                          list("q2m", 150, TAU_Q2M, 7L)), function(z) {
  tau <- z[[3]]; nd <- z[[4]]
  s <- simRegimen(z[[2]], tau, nd, grid = seq(0, nd * tau, by = 24))
  per <- s |>
    mutate(k = pmin(floor(time / tau) + 1L, nd)) |>
    group_by(id, k) |>
    arrange(time, .by_group = TRUE) |>
    summarise(auc = trapz(time, TAM), .groups = "drop")
  per |>
    left_join(per |> filter(k == nd) |> select(id, ssauc = auc), by = "id") |>
    group_by(k) |>
    summarise(pct = 100 * median(auc / ssauc), .groups = "drop") |>
    mutate(regimen = z[[1]], months = k * (tau / TAU_Q1M),
           dose = paste(z[[2]], "mg"))
}) |> bind_rows()

attainment |>
  filter(months %in% c(1, 2, 3, 4, 6, 8, 12)) |>
  transmute(Regimen = regimen, Dose = dose,
            `Months since initiation` = months,
            `Percent of steady-state AUC` = pct) |>
  knitr::kable(digits = 1,
               caption = "Median per-interval TAM AUC as a percent of the same participant's steady-state interval. The q2m intervals are 2 months long, so only even months exist for that regimen.")
```

```{r steady-state-check, echo = FALSE, results = "asis"}
six <- attainment |> filter(months == 6)
cat("At 6 months the simulated cohort has reached ",
    paste(sprintf("%.0f%% (%s)", six$pct, six$regimen), collapse = " and "),
    " of steady-state TAM exposure, against the paper's \"> 90%\".\n", sep = "")
```

The 6-month "> 90%" criterion is reproduced for both regimens. The 3-month
figures run a few points above the paper's 86% / 88%; that comparison is
definition-sensitive (for q2m, three months falls in the middle of a dosing
interval rather than at an interval boundary), so it is treated here as a
directional rather than a numerical check.

```{r ss-plot, echo = FALSE}
ggplot(attainment, aes(months, pct, colour = regimen)) +
  geom_hline(yintercept = 90, linetype = "dashed", colour = "grey40") +
  geom_line(linewidth = 0.8) + geom_point(size = 1.6) +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  labs(x = "Months since initiation (28-day months)",
       y = "Percent of steady-state TAM AUC", colour = "Regimen",
       title = "Approach to steady state",
       subtitle = "Dashed line: the paper's > 90% criterion, met by ~6 months") +
  theme_bw()
```

# Injection site

Replicates Figure 4 of Perlstein 2025 (impact of injection site on TAM exposure).
Upper-arm injection raises `ka1` by 33%, but the paper's conclusion is that this
does not translate into an exposure difference, "indicating that the sites of
injection can be interchangeable".

```{r inj-site}
siteSim <- lapply(c(0L, 1L), function(arm) {
  cov <- cohort |> mutate(INJSITE_ARM = arm)
  nd <- 13L
  grid <- seq((nd - 1) * TAU_Q1M, nd * TAU_Q1M, by = 6)
  simRegimen(75, TAU_Q1M, nd, grid, cov = cov) |>
    mutate(site = if (arm == 1L) "Upper arm" else "Abdomen")
}) |> bind_rows()

siteExposure <- siteSim |>
  group_by(site, id) |>
  arrange(time, .by_group = TRUE) |>
  summarise(auc = trapz(time, TAM), cmax = max(TAM),
            ctrough = TAM[which.max(time)], .groups = "drop") |>
  group_by(site) |>
  summarise(across(c(auc, cmax, ctrough), median), .groups = "drop")

siteExposure |>
  rename("Injection site" = site, "AUCss (ng*h/mL)" = auc,
         "Cmax,ss (ng/mL)" = cmax, "Ctrough,ss (ng/mL)" = ctrough) |>
  knitr::kable(digits = 1,
               caption = "Median steady-state TAM exposure, TV-46000 75 mg q1m")
```

```{r inj-site-plot, echo = FALSE}
siteSim |>
  mutate(days = (time - 12 * TAU_Q1M) / 24) |>
  group_by(site, days) |>
  summarise(p5 = quantile(TAM, 0.05), p50 = median(TAM),
            p95 = quantile(TAM, 0.95), .groups = "drop") |>
  ggplot(aes(days, p50, colour = site, fill = site)) +
  geom_ribbon(aes(ymin = p5, ymax = p95), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  labs(x = "Days since the steady-state dose", y = "TAM (ng/mL)",
       colour = "Injection site", fill = "Injection site",
       title = "Injection site has no material effect on steady-state exposure",
       subtitle = "TV-46000 75 mg q1m; replicates Figure 4 of Perlstein 2025") +
  theme_bw()
```

```{r inj-site-check, echo = FALSE, results = "asis"}
d <- 100 * (siteExposure$auc[siteExposure$site == "Upper arm"] /
              siteExposure$auc[siteExposure$site == "Abdomen"] - 1)
cat("Upper-arm versus abdominal steady-state AUC differs by ",
    sprintf("%+.2f%%", d),
    " on the same simulated participants, despite the 33% higher `ka1`. ",
    "The reason is structural rather than coincidental: at steady state every ",
    "milligram injected is eventually absorbed by one route or the other, so ",
    "`AUCss` is fixed at `1000 * D / CL` regardless of how the depot splits ",
    "between them. Shifting the split changes the *shape* of the profile, not ",
    "its area -- which is exactly the paper's interchangeability conclusion.\n",
    sep = "")
```

# Dopamine D2-receptor occupancy

Replicates Figure 5 of Perlstein 2025. The paper simulated D2RO from each TAM
profile with a literature Emax model (Emax = 100%, kd = 10.1 ng/mL) and reported
that occupancy was "generally within the therapeutic window of 60-80% during both
dosing intervals". `D2RO` is a derived output of the packaged model, so no extra
simulation is required.

```{r d2ro, echo = FALSE}
sims |>
  filter(time >= (n_dose - 1) * tau) |>
  mutate(days = (time - (n_dose - 1) * tau) / 24) |>
  group_by(regimen, treatment, dose, days) |>
  summarise(p50 = median(D2RO), .groups = "drop") |>
  ggplot(aes(days, p50, colour = factor(dose))) +
  geom_hline(yintercept = c(60, 80), linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = 40, linetype = "dotted", colour = "grey60") +
  geom_line(linewidth = 0.8) +
  facet_wrap(~regimen, scales = "free_x") +
  coord_cartesian(ylim = c(30, 95)) +
  labs(x = "Days since the steady-state dose", y = "D2 receptor occupancy (%)",
       colour = "Dose (mg)", title = "Simulated D2 receptor occupancy",
       subtitle = "Dashed lines: the 60-80% therapeutic window; replicates Figure 5 of Perlstein 2025") +
  theme_bw()
```

```{r d2ro-window, echo = FALSE}
sims |>
  filter(time >= (n_dose - 1) * tau) |>
  group_by(regimen, dose, days = (time - (n_dose - 1) * tau) / 24) |>
  summarise(p50 = median(D2RO), .groups = "drop") |>
  group_by(regimen, dose) |>
  summarise(`Time in 60-80% (%)` = 100 * mean(p50 >= 60 & p50 <= 80),
            `Median D2RO min (%)` = min(p50),
            `Median D2RO max (%)` = max(p50), .groups = "drop") |>
  rename("Regimen" = regimen, "Dose (mg)" = dose) |>
  knitr::kable(digits = 1,
               caption = "Fraction of the steady-state dosing interval with median D2RO inside the 60-80% window")
```

Consistent with the paper's Figure 5, the median occupancy sits inside or close
to the 60-80% band for most of each dosing interval, with the lowest strengths
running below the band late in the interval and the highest strengths above it
shortly after dosing.

# Assumptions and deviations

* **Covariate centring values are the published means, not the medians.** The
  paper's continuous-covariate equation centres on `COV_median`, but Table 2
  reports only means and SDs, and the medians appear nowhere in the article or
  its supplement. The model uses the Table 2 overall means -- **BMI 28.7 kg/m^2**
  and **injection volume 0.303 mL** -- as centring values. Both are on-disk
  published numbers for the same population, and the resulting error is small: a
  100 mg injection (0.278 mL) is mis-scaled by
  `(0.278/0.303)^-0.384 - 1` = +3.4% on `ka1`, and a BMI shift of 0.2 kg/m^2
  moves `ka1` by under 1%. A user who knows the true medians can override the two
  constants in `model()`.
* **The molecular-weight correction in TAM is applied; the two weights are not
  published.** Perlstein 2025 defines its endpoint in the Introduction as "the
  total active moiety (TAM; sum of concentrations of risperidone and its active
  metabolite [9-OH risperidone] corrected by molecular weight)" but never prints
  the weights. The model uses `410 / 426`, the rounded values for risperidone
  (410.49 g/mol) and 9-hydroxyrisperidone (426.49 g/mol) that are standard in the
  risperidone LAI literature and are already used by the upstream
  `Ivaturi_2017_RBP_7000` model for the same two molecules. The factor matters:
  because the metabolite carries roughly 71% of TAM, dropping it would raise every
  TAM reading by about 2.7%. Note that this correction applies **only to the TAM
  summation**, not to the parent-to-metabolite ODE flux -- Table 3's footnote
  states that `VMO` is `V/FRMET` with FRMET fixed to 1, so `CLMO` and `VMO` are
  apparent values fitted to the measured 9-OH-risperidone concentrations and
  already absorb any mass conversion inside the disposition model.
  This definition is in the Introduction, which the preprocessed `_trimmed.md`
  companion of the source PDF strips; it was recovered from the raw PDF.
* **Injection volume is derived from the milligram dose.** `DOSE_TV46000_ML` is
  computed as `dose_mg / 360`, using the suspension strength implied by the four
  dose/volume pairs the paper prints. A user with the actual administered volume
  should supply it directly.
* **IIV percentages read as lognormal %CV.** See the "Reading the IIV
  percentages" section: the choice is settled by which convention reproduces
  Table 3's published confidence intervals, not by assumption.
* **`KA1BMI1` upper confidence bound.** Table 3 prints the KA1BMI1 interval as
  "-1.1 (-1.460 to 0.745)". The upper bound has lost its minus sign in
  typesetting: the point estimate is -1.1 with %RSE 16.6, giving
  -1.1 +/- 1.96 x 0.183 = (-1.458, -0.742), and the bootstrap row prints the same
  pattern. Only the point estimate enters the model, so nothing downstream
  depends on this, but the interval as printed is not usable.
* **Covariates screened but not retained are documented, not encoded.** Age and
  sex reached statistical significance on the parent and sex on the metabolite,
  but all were removed as not clinically relevant and **no point estimates are
  reported for them**, so they cannot be reconstructed. They are recorded in the
  model file's `covariatesDataExcluded`. The reported "-1.7%" for sex on CLMO is
  a change in objective function, not a covariate coefficient.
* **Product presentation (vial versus prefilled syringe) is not carried.** It was
  screened as an extrinsic factor and had no effect on exposure. No canonical
  register column exists for a vial-versus-prefilled-syringe contrast (the
  nearest, `DEVICE_AI`, contrasts an autoinjector against a prefilled syringe),
  so registering one for a covariate with no estimated effect was not warranted;
  it is recorded in the model's `population$notes` instead.
* **The oral risperidone comparator is not part of this model.** Perlstein 2025
  simulated oral risperidone 1-16 mg/d from a separate published model
  (its reference 22) purely as a benchmark for Table 4 and Figure 3. That model
  belongs to another paper and is not extracted here; the library's oral
  risperidone entries are `Feng_2008_risperidone` and `Sherwin_2012_risperidone`.
* **D2RO parameters are literature constants, not fitted here.** `Emax` = 100%
  and `kd` = 10.1 ng/mL come from the paper's references 22 and 23 and are
  encoded with `fixed()`.
* **Virtual cohort distributions are approximations.** BMI is drawn from a normal
  distribution with the phase 3 mean and SD, truncated to the Table 2 range;
  injection site is assigned per subject at the RISE arm frequency (34.4%) rather
  than per injection. The published simulations bootstrapped 500 real phase 3
  covariate vectors; this vignette uses `r NSUB` synthetic subjects per arm to
  stay inside the package's simulation-size budget, which is the main source of
  the residual scatter in the Table 4 comparison.
* **Steady state is evaluated at the 13th q1m interval and the 7th q2m
  interval** (approximately 12 and 13 months), well past the ~6 months the paper
  reports as full attainment.
