---
title: "Elranatamab (Poels 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Elranatamab (Poels 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
set.seed(20250825)
```

```{r model-meta, include = FALSE}
mod_fn <- readModelDb("Poels_2025_elranatamab_qsp")
mod <- rxode2::rxode(mod_fn)
mod_meta <- mod_fn()

MW_ELRA <- 148500        # g/mol, ELREXFIO USPI Description section
CYCLE_H <- 28 * 24       # one treatment cycle = 28 days
PM_PER_NGML <- 185       # Poels 2025 sBCMA scale (Suppl Table 2 vs Fig 2b)

## Every simulation goes through the packaged event-table helper so that the
## Supplementary Eq 37 cytokine-exposure reset is always applied.
solve_qsp <- function(dose_time, dose_mg, obs_time, params = NULL, n_id = 1L,
                      cytokine_reset = TRUE) {
  ev <- Poels_2025_elranatamab_qsp_events(
    dose_time = dose_time, dose_mg = dose_mg, obs_time = obs_time,
    cytokine_reset = cytokine_reset
  )
  if (n_id > 1L) {
    ev <- do.call(rbind, lapply(seq_len(n_id), function(i) {
      e <- ev
      e$id <- i
      e
    }))
  }
  out <- rxode2::rxSolve(
    mod, ev, params = params, returnType = "data.frame",
    # rxode2's ODE -> linCmt auto-conversion is not valid for this 28-state model
    useLinCmt = FALSE
  )
  out <- as.data.frame(out)
  if (is.null(out$id)) out$id <- 1L         # rxSolve drops `id` for one subject
  out[!is.na(out$Cc), , drop = FALSE]
}

## Untreated run: no dose records at all.
solve_untreated <- function(obs_time, params = NULL) {
  ev <- data.frame(id = 1L, time = obs_time, amt = NA_real_,
                   evid = 0L, cmt = "central")
  out <- as.data.frame(rxode2::rxSolve(mod, ev, params = params,
                                       returnType = "data.frame",
                                       useLinCmt = FALSE))
  if (is.null(out$id)) out$id <- 1L
  out
}

pm_to_ugml <- function(pm) pm * 1e-12 * MW_ELRA * 1e3   # pM -> ug/mL

## MagnetisMM-3 Cohort A priming: 12 mg C1D1, 32 mg C1D4, 76 mg C1D8.
prime_time <- c(0, 72, 168)
prime_mg <- c(12, 32, 76)
qw_times <- function(from, to) seq(from, to, by = 168)

th <- mod$theta

## Terminal half-life implied by the linear two-compartment parameters alone,
## for comparison with the NCA half-life of the full QSP model.
linear_thalf_days <- local({
  kel <- exp(th[["lcl"]]) / exp(th[["lvc"]])
  k12 <- exp(th[["lq"]]) / exp(th[["lvc"]])
  k21 <- exp(th[["lq"]]) / exp(th[["lvp"]])
  s <- kel + k12 + k21
  beta <- (s - sqrt(s^2 - 4 * kel * k21)) / 2
  log(2) / beta / 24
})
```

* Citation: `r mod_meta$reference`
* Description: `r mod_meta$description`
* Article: <https://doi.org/10.1038/s41540-025-00585-z>

Poels et al. (2025) develop a quantitative systems pharmacology (QSP) model of
**elranatamab**, an approved BCMA x CD3 bispecific antibody (BsAb) for relapsed
or refractory multiple myeloma (RRMM), and use it to justify the 76 mg weekly
regimen and the step-down to every-2-weeks (Q2W) and every-4-weeks (Q4W) dosing
in persistent responders. The model extends the CD3-bispecific framework of
Betts et al. (2019) -- already in this library as
`modellib("Betts_2019_pf_06671008_qsp")` -- with soluble BCMA (sBCMA) as a drug
sink, myeloma paraprotein read-outs, and the cytokine-release framework of
Chen et al. (2019).

The shipped model carries all `r length(mod$state)` ordinary differential
equations of Supplementary Note 1 with the parameter values of Supplementary
Tables 2 and 3. The paper reports no residual-error model and no
between-subject random effects: it is a simulation-only QSP model in which
patient-to-patient heterogeneity is expressed by resampling nine structural
parameters (Supplementary Table 1) and five baseline states (Supplementary
Table 2). `propSd` is therefore `fixed(0)`.

# Population

```{r population, echo = FALSE}
pop <- mod_meta$population
tibble::tibble(Field = names(pop), Value = vapply(pop, paste, character(1), collapse = "; ")) |>
  knitr::kable()
```

Calibration used the phase 1 **MagnetisMM-1** (C1071001, NCT03269136) and the
registrational phase 2 **MagnetisMM-3** (C1071003, NCT04649359) studies. The
paper reports the per-arm sample sizes and dosing schedules (Supplementary
Table 4) but no age, weight, sex or race distributions for the calibration
cohorts, so those fields are absent from the model metadata rather than guessed.

# Model structure

The three anatomical compartments are the **central** (circulating)
compartment, a **peripheral** distribution compartment, and the **bone marrow**,
which serves as the tumour site of action. In the marrow the BsAb binds
membrane BCMA on myeloma cells and CD3 on T cells; the productive
BsAb-CD3-BCMA **trimer** drives myeloma-cell killing and cytokine release.
Soluble BCMA, shed in proportion to tumour burden, binds one arm of the BsAb in
both compartments to form non-productive dimers -- the drug-sink mechanism the
paper identifies as the key determinant of response.

```{r structure-table, echo = FALSE}
cd <- mod_meta$compartmentData
tibble::tibble(
  State = names(cd),
  Analyte = vapply(cd, function(x) x$analyte, character(1)),
  Units = vapply(cd, function(x) x$units, character(1)),
  Specimen = vapply(cd, function(x) x$specimen, character(1))
) |>
  knitr::kable(caption = "ODE states of the shipped model.")
```

## Dosing convention

`depot` holds a subcutaneous **amount in pmol**, because Supplementary Eq 1
writes the absorption flux as `F * ka * [Ab_sc] / Vc`; every other drug state
holds a **concentration in pM**. Doses are therefore converted from mg to pmol
with the elranatamab molecular weight of 148.5 kDa (ELREXFIO US prescribing
information, Description section), so 76 mg = `r format(round(76 / MW_ELRA * 1e9), big.mark = ",")` pmol.

`Poels_2025_elranatamab_qsp_events()` performs that conversion and, critically,
emits the Supplementary Eq 37 reset of the cumulative-cytokine-exposure state
as an `evid = 6` (multiply) record at each dose time. That reset is a discrete
state assignment which an rxode2 ODE cannot express, so it must come from the
event table; simulating without it silently removes the dose-to-dose
attenuation of cytokine release.

```{r events-demo}
Poels_2025_elranatamab_qsp_events(
  dose_time = c(0, 72, 168),
  dose_mg = c(12, 32, 76),
  obs_time = c(0, 24, 168)
)
```

# Source trace

Every `ini()` value carries an in-file comment naming its source location. The
table below is generated by parsing those comments out of the installed model
file, so it cannot drift from the model.

```{r source-trace-params, echo = FALSE}
model_path <- system.file("modeldb", "specificDrugs",
                          "Poels_2025_elranatamab_qsp.R",
                          package = "nlmixr2lib")
src <- readLines(model_path)
ini_start <- grep("^\\s*ini\\(\\{", src)[1]
ini_end <- grep("^\\s*\\}\\)\\s*$", src)
ini_end <- ini_end[ini_end > ini_start][1]
ini_lines <- src[seq(ini_start + 1L, ini_end - 1L)]
keep <- grepl("^\\s*[A-Za-z][A-Za-z0-9_]*\\s*<-", ini_lines) & grepl("#", ini_lines)
trace <- ini_lines[keep]
code <- sub("#.*$", "", trace)
param <- sub("^\\s*([A-Za-z][A-Za-z0-9_]*)\\s*<-.*$", "\\1", code)
value <- trimws(sub("^\\s*[A-Za-z][A-Za-z0-9_]*\\s*<-\\s*", "", code))
value <- trimws(sub(";\\s*label\\(.*$", "", value))
source_loc <- trimws(sub("^[^#]*#\\s*", "", trace))
tibble::tibble(Parameter = param, `ini() value` = value, `Source location` = source_loc) |>
  knitr::kable(caption = "Source trace for every ini() parameter, parsed from the installed model file.")
stopifnot(length(param) > 40L)   # a silently-empty parse would render an empty table
```

```{r source-trace-eqs, echo = FALSE}
tibble::tribble(
  ~`Model element`, ~`Source location`,
  "d/dt(central) - free BsAb in circulation", "Suppl Note 1 Eq 1",
  "d/dt(target) - free sBCMA in circulation", "Suppl Note 1 Eqs 2-3",
  "d/dt(complex) - BsAb-sBCMA dimer, central", "Suppl Note 1 Eq 4",
  "d/dt(drug_cd3_central) - BsAb-CD3 dimer, central", "Suppl Note 1 Eq 5",
  "d/dt(tcell_central) - circulating T cells", "Suppl Note 1 Eq 6",
  "d/dt(cytokine_central) - circulating cytokine", "Suppl Note 1 Eq 7",
  "CD3 receptor pool, central", "Suppl Note 1 Eqs 8-10",
  "d/dt(mprotein), d/dt(flc) - paraproteins", "Suppl Note 1 Eqs 11-12",
  "d/dt(bonemarrow) - free BsAb in bone marrow", "Suppl Note 1 Eq 13",
  "d/dt(target_bonemarrow) - free sBCMA in bone marrow", "Suppl Note 1 Eq 14",
  "d/dt(drug_bcma_bonemarrow) - BsAb-BCMA dimer", "Suppl Note 1 Eq 15",
  "d/dt(complex_bonemarrow) - BsAb-sBCMA dimer, marrow", "Suppl Note 1 Eqs 16-17",
  "d/dt(drug_cd3_bonemarrow) - BsAb-CD3 dimer, marrow", "Suppl Note 1 Eq 18",
  "d/dt(trimer) - BsAb-CD3-BCMA trimer", "Suppl Note 1 Eq 19",
  "BCMA and CD3 receptor pools, marrow", "Suppl Note 1 Eqs 20-23",
  "d/dt(cycling_cells), d/dt(damaged_cells1-3) - Simeoni growth and kill", "Suppl Note 1 Eqs 24-27",
  "KMkill - trimer-driven kill rate", "Suppl Note 1 Sect 1.2.2 (unnumbered)",
  "km_kill_t - time-dependent resistance", "Suppl Note 1 Sect 1.2.2 (unnumbered)",
  "d/dt(tcell_bonemarrow) - marrow T cells", "Suppl Note 1 Eq 28",
  "t_inh - T-cell egress inhibition", "Suppl Note 1 Eq 29",
  "d/dt(cytokine_bonemarrow), d/dt(cytokine_transit1-5), d/dt(cauc)", "Suppl Note 1 Eqs 30-36",
  "cauc reset at each dose (event-table evid = 6)", "Suppl Note 1 Eq 37",
  "IH and Rsyn - cytokine release inhibition and rate", "Suppl Note 1 Eq 38",
  "trimerFraction - standardised trimer proportion", "Suppl Note 1 Eq 39",
  "Initial conditions of every state", "Suppl Table 2",
  "sBCMA pM <-> ng/mL scale (185 pM per ng/mL)", "Suppl Table 2 range vs Fig 2b axis",
  "Elranatamab molecular weight 148.5 kDa", "ELREXFIO USPI, Description"
) |>
  knitr::kable(caption = "Source trace for every model equation.")
```

# The reference patient

The paper publishes **sampling ranges**, not point estimates, for the five
sampled baseline states: every published result comes from a 120-patient
virtual population. The shipped defaults are therefore the geometric midpoint
of each published range (Figure 2b shows all of these distributions are
strongly right-skewed, which rules out the arithmetic midpoint), and they are
not presented as published estimates. They are listed in Assumptions and
deviations below.

```{r reference-baselines, echo = FALSE}
ref <- tibble::tibble(
  Quantity = c("Free sBCMA, central", "CD3+ T cells, central",
               "Myeloma burden, marrow", "Serum M-protein", "Involved serum FLC"),
  Default = c(
    sprintf("%.0f pM (%.1f ng/mL)", exp(th[["lsbcma0_central"]]),
            exp(th[["lsbcma0_central"]]) / PM_PER_NGML),
    sprintf("%.0f cells/uL", exp(th[["ltcell0_central"]])),
    sprintf("%s cells/uL", format(round(exp(th[["lmm0"]])), big.mark = ",")),
    sprintf("%.1f g/L", exp(th[["lmprotein0"]])),
    sprintf("%.0f mg/L", exp(th[["lflc0"]]))
  ),
  `Published range (Suppl Table 2)` = c(
    "185-153520 pM (1-830 ng/mL)", "16-2580 cells/uL", "0-450000 cells/uL",
    "0-70 g/L", "0.92-174680 mg/L"
  )
)
knitr::kable(ref)
```

## Structural checks on the untreated system

Supplementary Sections 1.1.1 and 1.2.1 state that the system is at steady state
at `t = 0`. Three exact identities follow, and the untreated model must satisfy
all three.

```{r untreated}
untreated <- solve_untreated(seq(0, 3000, by = 10))
first_last <- untreated[c(1L, nrow(untreated)), ]

## (1) T cells are conserved at baseline in both compartments.
tcell_drift <- max(abs(untreated$tcell_central / untreated$tcell_central[1] - 1),
                   abs(untreated$tcell_bonemarrow / untreated$tcell_bonemarrow[1] - 1))

## (2) With no trimer there is no cytokine release, so circulating cytokine
##     relaxes to the basal balance beta_prod / kdeg_cyt.
cyt_ss_expected <- exp(th[["lbeta_prod"]]) / exp(th[["lkdeg_cyt"]])
cyt_ss_observed <- untreated$cytokineSerum[nrow(untreated)]

## (3) Untreated myeloma grows at kg0 while burden is far below MMmax, so the
##     doubling time approaches log(2) / kg0 exactly. Measured on a low-burden
##     patient (1% of the published baseline range) so the logistic term is
##     negligible and this is a true identity rather than an approximation.
doubling_time <- function(sim) {
  late <- sim[sim$time >= 1000, ]
  log(2) / coef(lm(log(late$tumorBurden) ~ late$time))[[2]] / 24
}
dt_expected_days <- log(2) / exp(th[["lkg0"]]) / 24
untreated_low <- solve_untreated(seq(0, 3000, by = 10),
                                 params = c(lmm0 = log(4500)))
dt_low_days <- doubling_time(untreated_low)
dt_ref_days <- doubling_time(untreated)
```

```{r untreated-table, echo = FALSE}
tibble::tibble(
  Check = c("Max relative drift in T cells (both compartments)",
            "Circulating cytokine at t = 3000 h (pg/mL)",
            "Doubling time, low-burden patient (days)",
            "Doubling time, reference patient (days)"),
  Model = c(sprintf("%.2e", tcell_drift),
            sprintf("%.4f", cyt_ss_observed),
            sprintf("%.1f", dt_low_days),
            sprintf("%.1f", dt_ref_days)),
  Expected = c("0 (exact conservation)",
               sprintf("%.4f  (= beta_prod / kdeg_cyt)", cyt_ss_expected),
               sprintf("%.1f  (= log(2) / kg0)", dt_expected_days),
               sprintf("> %.1f  (logistic term slows growth)", dt_expected_days))
) |>
  knitr::kable(caption = "Structural identities of the untreated system.")

stopifnot(
  tcell_drift < 1e-6,
  abs(cyt_ss_observed - cyt_ss_expected) < 1e-3,
  # at 1% of the baseline range the logistic term is negligible, so this is
  # an identity and is asserted to 1%
  abs(dt_low_days / dt_expected_days - 1) < 0.01,
  # at the reference burden the (1 - MMtot/MMmax) term must measurably slow growth
  dt_ref_days > dt_low_days * 1.05
)
```

Note that the baseline circulating cytokine value in Supplementary Table 2
(2.857 pg/mL) is **not** the model's own basal steady state
(`beta_prod / kdeg_cyt` = `r sprintf("%.2f", cyt_ss_expected)` pg/mL), so an
untreated simulation relaxes from the tabulated initial condition to
`r sprintf("%.2f", cyt_ss_expected)` pg/mL within a few hours. Both values are
transcribed exactly as published; the discrepancy is noted below.

# Pharmacokinetics

A single 76 mg subcutaneous dose, with the QSP binding and marrow-transport
terms active.

```{r pk-single}
pk <- solve_qsp(dose_time = 0, dose_mg = 76, obs_time = seq(0, 2016, by = 6))
pk$conc_ugml <- pm_to_ugml(pk$Cc)
```

```{r pk-plot, echo = FALSE, fig.cap = "Free elranatamab in serum after a single 76 mg SC dose."}
ggplot(pk, aes(time / 24, conc_ugml)) +
  geom_line(linewidth = 0.8, colour = "#2166ac") +
  labs(x = "Time (days)", y = expression(paste("Free elranatamab (", mu, "g/mL)"))) +
  theme_bw()
```

## NCA

The model applies bioavailability internally at the depot, so the NCA dose is
the **absorbed** amount; clearance from `Dose/AUC` is then comparable to the
model's `CL` rather than to `CL/F`.

```{r pknca-single}
f_sc <- exp(th[["lfdepot"]])
dose_absorbed_mg <- 76 * f_sc

nca_single <- pk |>
  dplyr::transmute(id = id, treatment = "76 mg SC single dose",
                   time = time, Cc = conc_ugml)
dose_single <- data.frame(id = 1L, treatment = "76 mg SC single dose",
                          time = 0, amt = dose_absorbed_mg)

nca_res <- suppressWarnings(as.data.frame(PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(nca_single, Cc ~ time | treatment + id,
                   concu = "ug/mL", timeu = "h"),
  PKNCA::PKNCAdose(dose_single, amt ~ time | treatment + id, doseu = "mg"),
  intervals = data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                         auclast = TRUE, half.life = TRUE)
))))
```

The label's clearance is a **steady-state** value ("0.324 L/day following
24 weeks dosing"), so the matching simulated quantity is `Dose/AUCtau` over the
last complete weekly interval of a 24-week weekly regimen -- not a single-dose
`AUCinf`, which over any tractable observation window would be dominated by
extrapolation.

```{r pknca-ss}
ss_end <- 24 * 168
ss_dose_t <- c(prime_time, qw_times(336, ss_end))
ss_dose_mg <- c(prime_mg, rep(76, length(ss_dose_t) - 3L))
ss <- solve_qsp(ss_dose_t, ss_dose_mg, seq(0, ss_end, by = 2))
ss$conc_ugml <- pm_to_ugml(ss$Cc)

last_dose_t <- max(ss_dose_t[ss_dose_t <= ss_end - 168])
nca_ss_res <- suppressWarnings(as.data.frame(PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(
    dplyr::transmute(ss, id = id, treatment = "76 mg QW, week 24",
                     time = time, Cc = conc_ugml),
    Cc ~ time | treatment + id, concu = "ug/mL", timeu = "h"),
  PKNCA::PKNCAdose(
    data.frame(id = 1L, treatment = "76 mg QW, week 24",
               time = last_dose_t, amt = dose_absorbed_mg),
    amt ~ time | treatment + id, doseu = "mg"),
  intervals = data.frame(start = last_dose_t, end = last_dose_t + 168,
                         auclast = TRUE, cmax = TRUE, cmin = TRUE)
))))
```

```{r pknca-table, echo = FALSE}
nca_val <- function(res, p) {
  v <- res$PPORRES[res$PPTESTCD == p]
  if (length(v) != 1L) stop("no unique NCA result for '", p, "'")
  v
}
auctau <- nca_val(nca_ss_res, "auclast")
cl_ss_lday <- dose_absorbed_mg / auctau * 24
paper_cl_lday <- exp(th[["lcl"]]) * 24
usp_cl_lday <- 0.324    # ELREXFIO USPI 12.3
usp_tmax_d <- 7         # ELREXFIO USPI 12.3, median (range 3-7)
usp_thalf_d <- 22       # ELREXFIO USPI 12.3, 64% CV

pk_cmp <- tibble::tibble(
  `NCA parameter` = c("Cmax, single dose (ug/mL)", "Tmax, single dose (days)",
                      "AUC0-last, single dose (ug*h/mL)", "t1/2 (days)",
                      "Cmax at steady state (ug/mL)", "Ctrough at steady state (ug/mL)",
                      "AUCtau at steady state (ug*h/mL)", "CL at steady state (L/day)"),
  Simulated = c(
    sprintf("%.2f", nca_val(nca_res, "cmax")),
    sprintf("%.1f", nca_val(nca_res, "tmax") / 24),
    sprintf("%.0f", nca_val(nca_res, "auclast")),
    sprintf("%.1f", nca_val(nca_res, "half.life") / 24),
    sprintf("%.1f", nca_val(nca_ss_res, "cmax")),
    sprintf("%.1f", nca_val(nca_ss_res, "cmin")),
    sprintf("%.0f", auctau),
    sprintf("%.3f", cl_ss_lday)
  ),
  Reference = c("-", sprintf("%.0f (median, range 3-7)", usp_tmax_d), "-",
                sprintf("%.0f (64%% CV)", usp_thalf_d), "-", "-", "-",
                sprintf("%.3f", usp_cl_lday)),
  Source = c("-", "ELREXFIO USPI 12.3", "-", "ELREXFIO USPI 12.3",
             "-", "-", "-", "ELREXFIO USPI 12.3")
)
knitr::kable(pk_cmp, caption = "Simulated NCA against the independent ELREXFIO US prescribing information. None of the reference values was used in building the model.")
```

All three label comparisons are genuine out-of-sample checks: the model's PK
parameters came from the paper's own preliminary popPK, and no ELREXFIO label
value entered the extraction except the molecular weight.

* **Tmax** is `r sprintf("%.1f", nca_val(nca_res, "tmax") / 24)` days against
  the label's median of 7 days. The trial's PK sampling was on nominal study
  days, so the label's "7" is a visit label rather than a resolved maximum.
* **Terminal half-life** is
  `r sprintf("%.1f", nca_val(nca_res, "half.life") / 24)` days against the
  label's 22 days (64% CV). Note this is *shorter* than the
  `r sprintf("%.0f", linear_thalf_days)`
  days the linear two-compartment parameters alone would give, because the QSP
  layer eliminates the BsAb-sBCMA dimer at the same rate as free drug and so
  adds a target-mediated component.
* **Steady-state clearance** is `r sprintf("%.3f", cl_ss_lday)` L/day against
  the label's 0.324 L/day, and the paper's own linear
  `CL = `r sprintf("%.3f", paper_cl_lday)`` L/day sits between them.

The one quantity that does not line up is volume: the label reports a
steady-state volume of 7.76 L against the model's
`Vc + Vp = `r sprintf("%.2f", exp(th[["lvc"]]) + exp(th[["lvp"]]))`` L. The
paper's PK parameters are explicitly labelled "Preliminary PopPK" -- an interim
fit available when the QSP work was done -- so a difference from the final
label popPK is a property of the published parameter set, not of the
transcription.

```{r pknca-assert, echo = FALSE}
stopifnot(
  # Tmax within a day of the label median
  abs(nca_val(nca_res, "tmax") / 24 - usp_tmax_d) <= 1,
  # terminal half-life within 2-fold of the label value
  nca_val(nca_res, "half.life") / 24 > usp_thalf_d / 2,
  nca_val(nca_res, "half.life") / 24 < usp_thalf_d * 2,
  # steady-state clearance within 2-fold of the label value
  cl_ss_lday > usp_cl_lday / 2, cl_ss_lday < usp_cl_lday * 2,
  # concentrations never decay into negative solver noise
  all(pk$Cc >= 0), all(ss$Cc >= 0)
)
```

# MagnetisMM-3 Cohort A regimen

The registrational regimen is 12 mg on C1D1, 32 mg on C1D4 and 76 mg on C1D8,
weekly thereafter (Supplementary Table 4).

```{r mm3-regimen}
mm3_end <- 6 * CYCLE_H
mm3_dose_t <- c(prime_time, qw_times(336, mm3_end))
mm3_dose_mg <- c(prime_mg, rep(76, length(mm3_dose_t) - 3L))
mm3 <- solve_qsp(mm3_dose_t, mm3_dose_mg, seq(0, mm3_end, by = 6))
```

```{r mm3-plot, echo = FALSE, fig.height = 6, fig.cap = "Reference-patient dynamics under the MagnetisMM-3 Cohort A regimen. Replicates the shape of Figure 2c (paraprotein), Supplementary Figure 7 (free sBCMA) and Supplementary Figure 8 (IL-6)."}
mm3 |>
  dplyr::transmute(
    time = time / 24,
    `Free elranatamab (ug/mL)` = pm_to_ugml(Cc),
    `Free sBCMA (ng/mL)` = sbcmaFree,
    `Serum cytokine / IL-6 (pg/mL)` = cytokineSerum,
    `M-protein (% change from baseline)` = 100 * (mProtein / mProtein[1] - 1)
  ) |>
  tidyr::pivot_longer(-time) |>
  ggplot(aes(time, value)) +
  geom_line(linewidth = 0.7, colour = "#b2182b") +
  facet_wrap(~name, scales = "free_y", ncol = 2) +
  labs(x = "Time (days)", y = NULL) +
  theme_bw()
```

## Cytokine-release attenuation (Supplementary Figure 8)

Supplementary Figure 8 shows IL-6 peaking during the priming doses and then
returning to baseline as repeated dosing inflates the cumulative-exposure state
and drives the inhibition term `IH` toward `Imax`. The model reproduces that
attenuation, which is what the Supplementary Eq 37 reset exists to produce.

```{r cytokine-peaks}
peak_after <- function(sim, t0) max(sim$cytokineSerum[sim$time >= t0 & sim$time <= t0 + 160])
cyt_peaks <- vapply(mm3_dose_t[1:8], function(t0) peak_after(mm3, t0), numeric(1))

## Same regimen with the Eq 37 reset suppressed.
mm3_noreset <- solve_qsp(mm3_dose_t, mm3_dose_mg, seq(0, mm3_end, by = 6),
                         cytokine_reset = FALSE)
cyt_peaks_noreset <- vapply(mm3_dose_t[1:8], function(t0) peak_after(mm3_noreset, t0), numeric(1))
```

```{r cytokine-table, echo = FALSE}
tibble::tibble(
  Dose = seq_along(cyt_peaks),
  `Day` = mm3_dose_t[1:8] / 24,
  `Peak cytokine, with Eq 37 reset (pg/mL)` = round(cyt_peaks, 2),
  `Peak cytokine, reset suppressed (pg/mL)` = round(cyt_peaks_noreset, 2)
) |>
  knitr::kable(caption = "Serum cytokine peak after each of the first eight doses.")

stopifnot(
  # peaks attenuate monotonically from the first priming dose onward
  all(diff(cyt_peaks) <= 1e-8),
  # and return to the basal balance by the sixth dose
  abs(cyt_peaks[6] - cyt_ss_expected) < 0.1,
  # the reset is what produces the attenuation: without it the peaks are higher
  cyt_peaks_noreset[4] > cyt_peaks[4]
)
```

Peak cytokine falls from `r sprintf("%.1f", cyt_peaks[1])` pg/mL after the
12 mg priming dose to `r sprintf("%.1f", cyt_peaks[4])` pg/mL by the fourth
dose and is back at the basal balance by the sixth. Suppressing the Eq 37
reset raises the fourth-dose peak to
`r sprintf("%.1f", cyt_peaks_noreset[4])` pg/mL, confirming the reset is
load-bearing rather than cosmetic.

# Bell-shaped dose-response (Figure 4)

The paper's central mechanistic claim is that trimer formation -- and hence
efficacy -- is **not monotone in dose**: at high concentrations the BsAb
saturates BCMA and CD3 separately, producing an excess of non-productive dimers
at the expense of trimers. Figure 4 plots the "effective binding ratio"
(trimers per BCMA receptor) against average drug concentration for the five
simulated weekly doses; Supplementary Figure 10 classifies each virtual patient
as increasing, bell-shaped or decreasing.

```{r bell}
sim_doses <- c(16, 28, 44, 76, 152)
bell_end <- 3 * CYCLE_H

binding_curve <- function(params = NULL) {
  purrr_rows <- lapply(sim_doses, function(d) {
    dt <- qw_times(0, bell_end)
    s <- solve_qsp(dt, rep(d, length(dt)), seq(0, bell_end, by = 6), params = params)
    data.frame(dose_mg = d, conc = mean(s$Cc), binding_ratio = mean(s$bindingRatio))
  })
  do.call(rbind, purrr_rows)
}

curves <- dplyr::bind_rows(
  cbind(phenotype = "Reference patient", binding_curve()),
  cbind(phenotype = "High T cells (2000/uL)",
        binding_curve(c(ltcell0_central = log(2000)))),
  cbind(phenotype = "Low T cells (30/uL)",
        binding_curve(c(ltcell0_central = log(30))))
)
```

```{r bell-plot, echo = FALSE, fig.cap = "Effective binding ratio (trimers per BCMA receptor) versus mean drug concentration over the first three cycles, for three T-cell phenotypes. Replicates Figure 4a."}
ggplot(curves, aes(pm_to_ugml(conc), binding_ratio, colour = phenotype)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  scale_x_log10() +
  labs(x = expression(paste("Mean free elranatamab (", mu, "g/mL)")),
       y = "Trimer : BCMA receptor ratio", colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r bell-table, echo = FALSE}
shape_of <- function(br) {
  i <- which.max(br)
  if (i == 1L) "decreasing" else if (i == length(br)) "increasing" else "bell-shaped"
}
bell_summary <- curves |>
  dplyr::group_by(phenotype) |>
  dplyr::summarise(
    `Peak dose (mg QW)` = sim_doses[which.max(binding_ratio)],
    `Curve shape` = shape_of(binding_ratio),
    .groups = "drop"
  )
knitr::kable(bell_summary, caption = "Dose-binding-ratio curve shape by phenotype (Supplementary Figure 10 classification).")

ref_br <- curves$binding_ratio[curves$phenotype == "Reference patient"]
stopifnot(
  # the reference patient is genuinely non-monotone: the maximum is interior
  which.max(ref_br) > 1L, which.max(ref_br) < length(ref_br),
  # and the model reproduces more than one of the three published curve shapes
  dplyr::n_distinct(bell_summary$`Curve shape`) >= 2L
)
```

The reference patient peaks at
`r sim_doses[which.max(ref_br)]` mg weekly and loses
`r sprintf("%.0f%%", 100 * (1 - min(ref_br) / max(ref_br)))` of its peak
binding ratio at 152 mg -- the bell shape of Figure 4a. The three published
curve types of Supplementary Figure 10 are recovered by varying baseline
T-cell count alone, matching the paper's finding that marrow T-cell
concentration is one of the two most predictive features of curve shape.

# Baseline sBCMA as a drug sink

The paper identifies **100 ng/mL** as the baseline sBCMA threshold that best
separates biochemical responders from non-responders (by iterative logistic
regression; Methods), and calibrates two distinct dose-response curves either
side of it (Figure 2a). Sweeping the reference patient's baseline sBCMA across
the whole published range reproduces a sharp efficacy cliff.

```{r sbcma-sweep}
sbcma_levels_ngml <- c(1, 10, 30, 60, 100, 150, 200, 400, 830)
sbcma_sweep <- lapply(sbcma_levels_ngml, function(ng) {
  s <- solve_qsp(mm3_dose_t, mm3_dose_mg, seq(0, mm3_end, by = 24),
                 params = c(lsbcma0_central = log(ng * PM_PER_NGML)))
  data.frame(
    sbcma_ngml = ng,
    mean_trimer = mean(s$trimer),
    best_pcfb = 100 * (min(s$mProtein) / s$mProtein[1] - 1)
  )
}) |>
  dplyr::bind_rows()
```

```{r sbcma-table, echo = FALSE}
sbcma_sweep |>
  dplyr::transmute(
    `Baseline sBCMA (ng/mL)` = sbcma_ngml,
    `Mean trimer (pM)` = signif(mean_trimer, 3),
    `Best M-protein change from baseline (%)` = round(best_pcfb, 1),
    `Biochemical response` = ifelse(best_pcfb <= -50, "yes", "no")
  ) |>
  knitr::kable(caption = "Baseline sBCMA sweep at 76 mg QW over six cycles, reference patient otherwise unchanged.")

resp_by_sbcma <- sbcma_sweep$best_pcfb <= -50
cliff_lo <- max(sbcma_sweep$sbcma_ngml[resp_by_sbcma])
cliff_hi <- min(sbcma_sweep$sbcma_ngml[!resp_by_sbcma & sbcma_sweep$sbcma_ngml > cliff_lo])
stopifnot(
  # response is retained at or below the published threshold and lost above it
  any(resp_by_sbcma), any(!resp_by_sbcma),
  cliff_lo >= 100, cliff_hi <= 200,
  # the sink works through trimer formation, not through some other route
  sbcma_sweep$mean_trimer[sbcma_sweep$sbcma_ngml == 830] <
    sbcma_sweep$mean_trimer[sbcma_sweep$sbcma_ngml == 100] / 10
)
```

For this patient the response is lost between
`r cliff_lo` and `r cliff_hi` ng/mL -- bracketing the 100 ng/mL cut-off the
authors derived independently from the clinical data -- and mean trimer collapses
by a factor of
`r sprintf("%.0f", sbcma_sweep$mean_trimer[sbcma_sweep$sbcma_ngml == 100] / sbcma_sweep$mean_trimer[sbcma_sweep$sbcma_ngml == 830])`
across the full sBCMA range.

Note the small **non-monotonicity below the cliff**: response improves slightly
from 1 to 100 ng/mL. That is the bell shape of the previous section, not a
transcription error -- a modest sBCMA sink lowers free drug and moves a patient
who is past the trimer peak back toward it. The drug-sink effect dominates only
once sBCMA is large enough to strip trimer formation outright.

# Virtual population

The paper builds each virtual population in two steps: sample 10,000
parametrizations, filter to **plausible patients** on untreated tumour
doubling time, then use a genetic algorithm to select 120 **virtual patients**
whose summary statistics match the observed biochemical response rates and
paraprotein trajectories. The second step cannot be reproduced here: its
objective-function targets (the per-arm observed BRRs and median trajectories)
appear only in figures, and the plausibility bounds on doubling time are not
published. What follows is therefore the **plausible population**, sampled
exactly as Supplementary Table 1 describes, and its absolute response rates are
expected to sit below the published ones.

```{r vpop}
n_vp <- 120   # the paper's virtual-population size

runif_fold <- function(n, v, fold) stats::runif(n, v / fold, v * fold)
runif_pct <- function(n, v, pct) stats::runif(n, v * (1 - pct), v * (1 + pct))
runif_log <- function(n, lo, hi) stats::runif(n, log(lo), log(hi))

vp_params <- data.frame(
  id = seq_len(n_vp),
  ## Nine perturbed structural parameters, Supplementary Table 1.
  ## Sampled uniformly on the linear scale, as the paper states.
  lcd3_density  = log(runif_pct(n_vp, 6e4, 0.25)),
  lbcma_density = log(runif_pct(n_vp, 1.259e4, 0.25)),
  lbeta_resis   = log(runif_fold(n_vp, 0.1, 2)),
  lalpha_resis  = log(runif_fold(n_vp, 0.5, 2)),
  lkg0          = log(runif_fold(n_vp, 3.32e-4, 10)),
  lkg1          = log(runif_fold(n_vp, 1500, 10)),
  lalpha_kill   = log(runif_fold(n_vp, 0.02, 10)),
  ltau_mm       = log(stats::runif(n_vp, 24, 30)),
  ln_kill       = log(stats::runif(n_vp, 0.5, 1.5)),
  ## Five sampled baseline states, Supplementary Table 2, drawn LOG-uniformly
  ## over the published ranges (see Assumptions and deviations).
  lsbcma0_central = runif_log(n_vp, 185, 153520),
  ltcell0_central = runif_log(n_vp, 16, 2580),
  lmm0            = runif_log(n_vp, 450, 450000),
  lmprotein0      = runif_log(n_vp, 5, 70),
  lflc0           = runif_log(n_vp, 0.92, 174680)
)

vp <- solve_qsp(mm3_dose_t, mm3_dose_mg, seq(0, mm3_end, by = 24),
                params = vp_params, n_id = n_vp)

## Biochemical response, per the paper's definition: a >= 50% decrease from
## baseline in the integrated paraprotein, sustained over two consecutive
## (3-weekly) tumour assessments.
assess_h <- seq(0, mm3_end, by = 21 * 24)
vp_resp <- vp |>
  dplyr::filter(time %in% assess_h) |>
  dplyr::group_by(id) |>
  dplyr::arrange(time, .by_group = TRUE) |>
  dplyr::summarise(
    baseline_sbcma = dplyr::first(sbcmaFree),
    baseline_mp = dplyr::first(mProtein),
    pcfb_best = 100 * (min(mProtein) / dplyr::first(mProtein) - 1),
    responder = {
      hit <- mProtein / dplyr::first(mProtein) <= 0.5
      any(hit & dplyr::lead(hit, default = FALSE))
    },
    .groups = "drop"
  ) |>
  dplyr::mutate(sbcma_group = ifelse(baseline_sbcma >= 100,
                                     "high (>= 100 ng/mL)", "low (< 100 ng/mL)"))
```

```{r vpop-table, echo = FALSE}
vp_brr <- vp_resp |>
  dplyr::group_by(`Baseline sBCMA` = sbcma_group) |>
  dplyr::summarise(
    `Virtual patients` = dplyr::n(),
    `Share of population (%)` = round(100 * dplyr::n() / n_vp, 1),
    `Biochemical response rate (%)` = round(100 * mean(responder), 1),
    `Median best change from baseline (%)` = round(stats::median(pcfb_best), 1),
    .groups = "drop"
  )
knitr::kable(vp_brr, caption = "Plausible-population biochemical response at 76 mg QW, stratified by baseline sBCMA (compare Figure 3a).")

brr_low <- vp_brr$`Biochemical response rate (%)`[vp_brr$`Baseline sBCMA` == "low (< 100 ng/mL)"]
brr_high <- vp_brr$`Biochemical response rate (%)`[vp_brr$`Baseline sBCMA` == "high (>= 100 ng/mL)"]
share_low <- vp_brr$`Share of population (%)`[vp_brr$`Baseline sBCMA` == "low (< 100 ng/mL)"]

stopifnot(
  nrow(vp_brr) == 2L,                       # both strata must be populated
  length(brr_low) == 1L, length(brr_high) == 1L,
  brr_low > brr_high,                       # Figure 3a ordering
  # log-uniform baseline sBCMA reproduces the paper's ~70/30 low/high split
  share_low > 60, share_low < 80
)
```

Two published quantities are recovered. First, the **low/high sBCMA split**:
`r sprintf("%.0f%%", share_low)` of the plausible population has baseline sBCMA
below 100 ng/mL, against the paper's "70% of VPop". This is a check on the
sampling distribution rather than on the model, and it is why the baseline
states are drawn log-uniformly (a linear-uniform draw over the same range puts
only ~12% below the threshold, which cannot be what the authors did). Second,
the **ordering of response rates**: BRR is
`r sprintf("%.0f%%", brr_low)` in the low-sBCMA stratum against
`r sprintf("%.0f%%", brr_high)` in the high stratum, the direction of Figure 3a.

The absolute rates are below the published `~80%` and `~38%` at 76 mg QW,
exactly as expected: these are plausible patients, not the genetic-algorithm
selected virtual patients that were chosen *because* their summary statistics
matched the trial.

```{r vpop-plot, echo = FALSE, fig.cap = "Best M-protein change from baseline for each plausible patient at 76 mg QW, ranked. Compare the waterfall of Figure 2d."}
vp_resp |>
  dplyr::arrange(dplyr::desc(pcfb_best)) |>
  dplyr::mutate(rank = dplyr::row_number()) |>
  ggplot(aes(rank, pmin(pcfb_best, 150), fill = sbcma_group)) +
  geom_col(width = 0.9) +
  geom_hline(yintercept = -50, linetype = 2) +
  scale_fill_manual(values = c("high (>= 100 ng/mL)" = "#d6604d",
                               "low (< 100 ng/mL)" = "#4393c3")) +
  labs(x = "Virtual patient (ranked)",
       y = "Best M-protein change from baseline (%, capped at 150)",
       fill = "Baseline sBCMA") +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Dose reduction after response (Figure 5)

The paper's most counter-intuitive result is that stepping **down** from weekly
to Q2W (and then Q4W) dosing in responders *improves* maintenance of response,
because lowering drug exposure frees CD3 receptors from non-productive
BsAb-CD3 dimers and lets more trimers form.

```{r stepdown}
horizon <- 18 * CYCLE_H
switch_q2w <- 6 * CYCLE_H     # start of cycle 7
switch_q4w <- 12 * CYCLE_H    # start of cycle 13

sched_qw <- c(prime_time, qw_times(336, horizon))
sched_q2w <- c(prime_time, qw_times(336, switch_q2w - 168),
               seq(switch_q2w, horizon, by = 336))
sched_q4w <- c(prime_time, qw_times(336, switch_q2w - 168),
               seq(switch_q2w, switch_q4w - 336, by = 336),
               seq(switch_q4w, horizon, by = 672))

obs_grid <- seq(0, horizon, by = 6)
step <- dplyr::bind_rows(
  cbind(regimen = "QW throughout",
        solve_qsp(sched_qw, c(prime_mg, rep(76, length(sched_qw) - 3L)), obs_grid)),
  cbind(regimen = "QW -> Q2W at C7",
        solve_qsp(sched_q2w, c(prime_mg, rep(76, length(sched_q2w) - 3L)), obs_grid)),
  cbind(regimen = "QW -> Q2W at C7 -> Q4W at C13",
        solve_qsp(sched_q4w, c(prime_mg, rep(76, length(sched_q4w) - 3L)), obs_grid))
)
```

```{r stepdown-plot, echo = FALSE, fig.height = 5.5, fig.cap = "Effect of dose-frequency reduction. Replicates Figure 5a (tumour burden), Figure 5c (trimer per tumour cell) and Supplementary Figure 5a (BsAb-CD3 dimer per T cell)."}
step |>
  dplyr::transmute(
    regimen, time = time / 24,
    `Tumour burden (cells/uL)` = tumorBurden,
    `Trimer : tumour cell` = trimer / tumorBurden,
    `BsAb-CD3 dimer : marrow T cell` = drug_cd3_bonemarrow / tcell_bonemarrow
  ) |>
  tidyr::pivot_longer(-c(regimen, time)) |>
  ggplot(aes(time, value, colour = regimen)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = c(switch_q2w, switch_q4w) / 24, linetype = 3) +
  facet_wrap(~name, scales = "free_y", ncol = 1) +
  labs(x = "Time (days)", y = NULL, colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r stepdown-table, echo = FALSE}
after_switch <- step |> dplyr::filter(time >= switch_q2w)
step_summary <- after_switch |>
  dplyr::group_by(Regimen = regimen) |>
  dplyr::summarise(
    `Mean trimer : tumour cell` = signif(mean(trimer / tumorBurden), 3),
    `Mean BsAb-CD3 dimer : T cell` = signif(mean(drug_cd3_bonemarrow / tcell_bonemarrow), 3),
    .groups = "drop"
  )
burden_at <- function(reg, tt) {
  s <- step[step$regimen == reg, ]
  s$tumorBurden[which.min(abs(s$time - tt))]
}
step_summary$`Tumour burden at C13` <-
  signif(vapply(step_summary$Regimen, burden_at, numeric(1), tt = switch_q4w), 4)
step_summary$`Tumour burden at C18` <-
  signif(vapply(step_summary$Regimen, burden_at, numeric(1), tt = horizon), 4)
knitr::kable(step_summary, caption = "Regimen comparison over cycles 7-18, reference patient.")

g <- function(col, reg) step_summary[[col]][step_summary$Regimen == reg]
stopifnot(
  # Figure 5c: dose reduction RAISES the trimer per tumour cell
  g("Mean trimer : tumour cell", "QW -> Q2W at C7") >
    g("Mean trimer : tumour cell", "QW throughout"),
  g("Mean trimer : tumour cell", "QW -> Q2W at C7 -> Q4W at C13") >
    g("Mean trimer : tumour cell", "QW throughout"),
  # Supplementary Figure 5a: and LOWERS the BsAb-CD3 dimer per T cell
  g("Mean BsAb-CD3 dimer : T cell", "QW -> Q2W at C7") <
    g("Mean BsAb-CD3 dimer : T cell", "QW throughout"),
  # Figure 5a: greater tumour shrinkage under the reduced-frequency regimens
  g("Tumour burden at C13", "QW -> Q2W at C7") <
    g("Tumour burden at C13", "QW throughout"),
  g("Tumour burden at C18", "QW -> Q2W at C7 -> Q4W at C13") <
    g("Tumour burden at C18", "QW throughout")
)
```

Switching to Q2W at cycle 7 raises the mean trimer-per-tumour-cell ratio over
cycles 7-18 from
`r format(g("Mean trimer : tumour cell", "QW throughout"), scientific = TRUE, digits = 3)`
to
`r format(g("Mean trimer : tumour cell", "QW -> Q2W at C7"), scientific = TRUE, digits = 3)`
while lowering the BsAb-CD3 dimer per marrow T cell from
`r sprintf("%.3f", g("Mean BsAb-CD3 dimer : T cell", "QW throughout"))` to
`r sprintf("%.3f", g("Mean BsAb-CD3 dimer : T cell", "QW -> Q2W at C7"))` -- the
exact mechanism the paper proposes, and reproduced here from the equations
alone. Tumour burden at cycle 13 is
`r sprintf("%.0f%%", 100 * (1 - g("Tumour burden at C13", "QW -> Q2W at C7") / g("Tumour burden at C13", "QW throughout")))`
lower under the step-down regimen.

# Claims checked

```{r claims, echo = FALSE}
tibble::tribble(
  ~Claim, ~`Source`, ~`Checked by`,
  "T cells hold their baseline exactly in the untreated system", "Suppl Sects 1.1.1, 1.2.1, Eqs 6 and 28", "Max relative drift < 1e-6 in both compartments (asserted)",
  "Untreated cytokine relaxes to the basal balance beta_prod/kdeg_cyt", "Suppl Table 3, Eq 7", "Agreement to < 1e-3 pg/mL (asserted)",
  "Untreated doubling time equals log(2)/kg0 at low burden", "Suppl Table 3, Eq 24", "Fitted slope within 1% (asserted)",
  "The logistic term measurably slows growth at higher burden", "Suppl Eq 24", "Reference-patient doubling time > 1.05x the low-burden value (asserted)",
  "Single-dose Tmax consistent with the label", "ELREXFIO USPI 12.3", "Simulated Tmax within 1 day of the 7-day median (asserted)",
  "Terminal half-life consistent with the label", "ELREXFIO USPI 12.3", "NCA t1/2 within 2-fold of 22 days (asserted)",
  "Steady-state clearance consistent with the label", "ELREXFIO USPI 12.3", "Dose/AUCtau at week 24 within 2-fold of 0.324 L/day (asserted)",
  "Simulated concentrations never go negative", "numerical hygiene", "all(Cc >= 0) for single-dose and steady-state runs (asserted)",
  "Cytokine peaks attenuate over repeated doses", "Suppl Fig 8, Eq 37", "Peaks monotone non-increasing; basal by dose 6 (asserted)",
  "The Eq 37 reset is what causes the attenuation", "Suppl Eq 37", "Peaks higher with the reset suppressed (asserted)",
  "Trimer formation is non-monotone in dose (bell shape)", "Fig 4a, Suppl Fig 10", "Interior maximum for the reference patient (asserted)",
  "More than one dose-response curve shape arises", "Suppl Fig 10a", ">= 2 distinct shapes across T-cell phenotypes (asserted)",
  "Response is lost above ~100 ng/mL baseline sBCMA", "Fig 2a, Methods", "Highest responding level >= 100 and lowest failing level <= 200 ng/mL (asserted)",
  "sBCMA acts through trimer suppression", "Fig 3a, Discussion", "Mean trimer falls > 10-fold across the range (asserted)",
  "~70% of the virtual population has low baseline sBCMA", "Results, Fig 3a", "Share between 60% and 80% (asserted)",
  "BRR is higher in the low-sBCMA stratum", "Fig 3a", "brr_low > brr_high (asserted)",
  "QW -> Q2W raises the trimer:tumour-cell ratio", "Fig 5c", "Mean ratio higher after switch (asserted)",
  "QW -> Q2W lowers the BsAb-CD3 dimer:T-cell ratio", "Suppl Fig 5a", "Mean ratio lower after switch (asserted)",
  "Dose reduction gives greater tumour shrinkage", "Fig 5a", "Lower burden at C13 and C18 (asserted)"
) |>
  knitr::kable()
```

Every row marked "asserted" is backed by a `stopifnot()` in the chunk that
computes it, so this vignette fails to render if any of them regresses.

Two published results are **not** reproduced here, and should not be read into
the tables above: the absolute biochemical response rates (which depend on the
genetic-algorithm virtual-patient selection, see above), and the simulated
progressive-disease rates of Figure 5b (which require the same selection step
plus the paper's persistent-responder enrolment logic).

# Assumptions and deviations

## Values not published, chosen here

* **Baseline state defaults.** Supplementary Table 2 publishes sampling ranges,
  not point estimates, for baseline sBCMA, T cells, myeloma burden, M-protein
  and FLC -- the paper has no "typical" patient. The shipped defaults are the
  **geometric midpoint** of each published range (Figure 2b confirms all are
  strongly right-skewed), except myeloma burden, whose range starts at zero and
  therefore uses the arithmetic midpoint, and M-protein, whose geometric
  midpoint is taken over 5-70 g/L using the IMWG measurability floor of
  0.5 g/dL that the paper itself cites. None of these is a published estimate.
* **Numerical stabiliser `eps_kill`.** Supplementary Section 1.2.2 introduces
  `epsilon` as "a small value introduced for numerical stability" in the
  trimer:myeloma-cell ratio and gives no value. It is a solver guard, not a
  calibrated quantity; the default 1e-6 cells/uL is twelve orders of magnitude
  below the smallest baseline burden in the sampling range.
* **Log-uniform baseline sampling in the virtual population.** The paper states
  the nine *parameters* of Supplementary Table 1 were sampled uniformly, and
  this vignette follows that exactly. For the *baseline states* it says only
  that distributions were "fitted to Pfizer baseline clinical data"; those
  distributions are not published. Figure 2b shows them strongly right-skewed,
  and drawing them log-uniformly over the Table 2 ranges reproduces the paper's
  stated ~70%/30% low/high sBCMA split (a linear-uniform draw gives ~12%/88%).

## Published values that conflict with each other

* **Myeloma-cell transit time.** Supplementary Table 1 gives a default of
  27 h with a sampling range of 24-30 h; Supplementary Table 3 gives 24 h. Every
  other Table 1 default agrees exactly with Table 3, and Table 1's 27 h is the
  midpoint of its own range. The model uses **Table 3's 24 h** as the
  authoritative full-parameter-table value; the virtual population samples
  24-30 h per Table 1.
* **Baseline circulating cytokine.** Supplementary Table 2 initialises the
  central cytokine state at 2.857 pg/mL, but the model's own basal balance is
  `beta_prod / kdeg_cyt` = `r sprintf("%.2f", cyt_ss_expected)` pg/mL. Both are
  transcribed as published; an untreated simulation relaxes from one to the
  other within a few hours.
* **M-protein units in Figure 2b.** Figure 2b labels the M-protein axis
  "(g/dL)" over a 0-65 range, but Supplementary Table 2 gives M-protein in g/L
  with a 0-70 range, and the paper elsewhere quotes the IMWG measurability
  threshold as 0.5 g/dL (= 5 g/L). 65 g/dL of M-protein is not physically
  possible, so the figure axis label is in error and the model uses **g/L**.
* **Mixed units in Supplementary Table 3.** Clearance is reported in mL/h
  (13.9583) while inter-compartmental clearance on the adjacent row is in L/h
  (0.008416). Each row is transcribed on the units printed for that row; the
  resulting CL of `r sprintf("%.3f", paper_cl_lday)` L/day cross-checks
  against the independent label value of 0.324 L/day, which confirms the
  reading.

## Equations transcribed with a documented correction or caveat

* **Resistance onset (Supplementary Section 1.2.2).** As printed,
  `km_kill = km * (1 + alpha_resis * (1 - exp(-beta_resis * (t - tau_resis))))`
  diverges to large *negative* half-maximal kill values before
  `t = tau_resis` (`exp(+0.1 * 600)` is of order 1e26), which contradicts the
  stated meaning of `tau_resis` as "the time it takes for the resistance to
  occur". The elapsed time is floored at zero, so resistance is absent until
  `tau_resis` and then rises to `alpha_resis`. This is the only equation
  changed from its printed form.
* **Mass balance of the marrow BsAb-sBCMA dimer.** Supplementary Eq 17 places
  the factor `V_BM / V_c` on the *loss* of `complex_bonemarrow` as well as on its
  *gain* in the central compartment (Eq 4); the free-drug pair (Eqs 1 and 13)
  carries that factor on the gain term only. The printed dimer pair is
  therefore not mass-conserving. It is transcribed exactly as printed.
* **Steady-state marrow baselines.** The supplement asserts twice that the
  system is at steady state at `t = 0`, but Supplementary Table 2 lists the
  marrow T-cell and marrow sBCMA states as independently sampled over the same
  ranges as their central counterparts. Their ODEs (Eqs 14 and 28) are
  stationary at `t = 0` only when the marrow value equals the central value
  times the transport ratio, so the model derives the two marrow baselines from
  the steady-state relations. Initialising them independently would leave the
  untreated system with an immediate transient in both, breaking the paper's own
  premise. With the derived values, T cells hold their baseline exactly and
  indefinitely (their dynamics are independent of tumour burden), while sBCMA is
  stationary only at `t = 0` and thereafter rises with the growing untreated
  tumour, as Supplementary Eq 14 requires.
* **Peripheral and subcutaneous ODEs.** Supplementary Note 1 writes out the
  central-compartment equation (Eq 1) but not `d/dt(Ab_sc)` or `d/dt(Ab_p)`.
  Both follow uniquely from Eq 1's own terms and the tabulated `ka`, `F`, `Q`,
  `Vc`, `Vp`: `k12 = Q/Vc` and `k21 = Q/Vp` reproduce Eq 1's
  `k21 * (Vp/Vc) * [Ab_p]` term exactly.
* **Dose-counter reset delivered from the event table.** Supplementary Eq 37 is
  a discrete reassignment of `cauc` at each dose, which no rxode2 ODE
  can express. It is emitted as an `evid = 6` multiply record by
  `Poels_2025_elranatamab_qsp_events()`. Computing the multiplier there rather
  than in `model()` keeps it independent of rxode2's internal ordering of
  same-time dose records; `Poels_2025_elranatamab_qsp_cauc_multiplier()` exposes
  the formula and is unit-tested against the printed equation.

## Tabulated but unused

Supplementary Table 3 lists a cytokine-release **priming factor `K` = 2.83**
and a **dose counter `N` = 1** sourced to Chen et al. (2019), but neither symbol
appears in any printed equation of Supplementary Note 1. They are not carried in
`ini()`, because an unreferenced parameter would be a convention violation, and
are recorded here instead.

## Encoding choices

* `Imax_TC` is listed as "Estimated" in Supplementary Table 3 with a value of
  exactly 1 -- the upper bound of a maximum-inhibition fraction. It is encoded as
  `fixed(1)` rather than log-transformed, so a refit cannot push it above 1.
* Parameters the paper reports as estimated or calibrated are log-transformed so
  a refit stays positive; parameters fixed from nonclinical data, literature or
  the upstream preliminary popPK are carried inside `fixed()`.
* The paper reports no residual-error model and no between-subject random
  effects, so `propSd` is `fixed(0)` and the model declares no `eta` terms.
