---
title: "Perioperative factor VIII (Preijers 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Perioperative factor VIII (Preijers 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Preijers T, Liesner R, Hazendonk HCAM, Chowdary P, Driessens MHE,
  Hart DP, Laros-van Gorkom BAP, van der Meer FJM, Meijer K, Fijnvandraat K,
  Leebeek FWG, Mathot RAA, Cnossen MH; OPTI-CLOT study group. Validation of a
  perioperative population factor VIII pharmacokinetic model with a large
  cohort of pediatric hemophilia A patients. *Br J Clin Pharmacol.*
  2021;87(11):4408-4420.
  doi:[10.1111/bcp.14864](https://doi.org/10.1111/bcp.14864). PMID:33884664.
* Description: two-compartment population PK model for perioperative factor
  VIII (FVIII) concentrates in children and adults with hemophilia A. The paper
  first externally validated the Hazendonk 2016 perioperative model
  (`Hazendonk_2016_factor_viii` in this package) on 87 children from Great
  Ormond Street Hospital (GOSH), found that it under-predicted their FVIII
  levels, and then re-estimated the model on the pooled original and new data.
  The re-estimated final model is packaged here.
* Modality: recombinant (88% of procedures) and plasma-derived FVIII
  concentrates given by intravenous bolus or continuous infusion; FVIII
  activity measured by one-stage clotting assay in IU/mL.

The final model (Preijers 2021 Table 2 footnote) is

\[
\begin{aligned}
\mathrm{CL}\,(\mathrm{mL/h}) &= 171 \cdot (\mathrm{BW}/68)^{0.75}
   \cdot (\mathrm{AGE}/40)^{-0.12} \cdot 1.14^{\mathrm{BG}} \\
V_{1}\,(\mathrm{mL}) &= 2930 \cdot (\mathrm{BW}/68)^{1.0}
   \cdot (\mathrm{AGE}/40)^{-0.09} \\
Q_{2}\,(\mathrm{mL/h}) &= 172 \cdot (\mathrm{BW}/68)^{0.75}, \qquad
V_{2}\,(\mathrm{mL}) = 1810 \cdot (\mathrm{BW}/68)
\end{aligned}
\]

with `BG = 1` for blood group O. The measured FVIII level is the model
prediction plus the patient's endogenous baseline level, reduced by 30% for
samples from patients on the B-domain-deleted product Refacto AF (Methods
Eq. 1; see Assumptions and deviations for how that equation is read).

# Population

The final model was fitted to **206 patients undergoing 342 surgical
procedures** (Preijers 2021 Table 1):

* the **original cohort** of Hazendonk 2016: 119 patients (children and
  adults) from five Dutch hemophilia treatment centres, median age 39.6 years
  (0.24-77.6), median weight 75 kg (5-111), 70% severe hemophilia A, 197
  procedures of which 49% were major surgery and 59% used continuous infusion;
* the **new cohort**: 87 children from GOSH, London, median age 2.57 years
  (0.03-15.2), median weight 14 kg (4-57), all with severe hemophilia A, 145
  minor procedures to insert, replace or remove a central venous access
  device, all dosed by bolus.

The total cohort had a median age of 7.79 years (0.03-77.6) and median weight
of 30 kg (4-111); 83% had severe hemophilia A and 39% blood group O (of 175
with a recorded blood group). Recombinant FVIII was used in 88% of procedures.
There were 2092 FVIII measurements (one-stage clotting assay), none below the
0.01 IU/mL quantification limit.

The same information is available programmatically via
`readModelDb("Preijers_2021_factor_viii")()$population`.

# Source trace

| Model element | Value | Source location |
|---------------|-------|-----------------|
| `lcl` (CL at 68 kg, 40 y, non-O) | log(0.171 L/h) | Table 2 final model: 171 mL/h/68 kg (RSE 7%) |
| `lvc` (V1 at 68 kg, 40 y) | log(2.930 L) | Table 2 final model: 2930 mL/68 kg (RSE 4%) |
| `lq` (Q2 at 68 kg) | log(0.172 L/h) | Table 2 final model: 172 mL/h/68 kg (RSE 19%) |
| `lvp` (V2 at 68 kg) | log(1.810 L) | Table 2 final model: 1810 mL/68 kg (RSE 10%) |
| `theta_bdp` (Refacto AF fraction) | 0.30 | Table 2 final model: 'B-domain deleted recombinant factor VIII' 0.30 (RSE 14%) |
| `e_age_cl` | -0.12 | Table 2 final model 'CL - Age' (RSE 26%); footnote `(AGE/40)^-0.12` |
| `e_age_vc` | -0.09 | Table 2 final model 'V1 - Age' (RSE 24%); footnote `(AGE/40)^-0.09` |
| `e_blood_cl` | 0.14 | Table 2 final model 'CL - Blood group O' 14% (RSE 6%); footnote `1.14^BG` |
| `e_wt_cl`, `e_wt_vc` (fixed) | 0.75, 1.0 | Methods after Eq. 2: exponents fixed to 0.75 (CL, Q2) and 1 (V1, V2) |
| `etalcl`, `etalvc` block | 0.14568, 0.05833, 0.07290 | Table 2 final model: IIV CL 39.6 %CV, IIV V1 27.5 %CV, correlation 56.6%; omega^2 = log(1 + CV^2), cov = r sqrt(omega^2_CL omega^2_V1) |
| `addSd_center123` / `_center45` / `_center6` | 0.12 / 0.06 / 0.17 IU/mL | Table 2 final model: additive SD by centre |
| `propSd_center123` / `_center45` / `_center6` | 0.197 / 0.21 / 0.22 | Table 2 final model: proportional %CV by centre |
| Two-compartment structure, 68 kg normalisation | -- | Results 3.3; Methods Eq. 2 |
| Age power model | -- | Methods Eq. 5; Results 3.3 (power form performed best) |
| Blood group O on CL | -- | Methods Eq. 3; Table 2 footnote |
| Baseline + Refacto AF observation equation | `Cc = (central/vc/1000 + FVIII_BL) * (1 - 0.30 * FORM_FVIII_BDD)` | Methods Eq. 1 |

Covariates tested but not retained: major surgery on CL (significant in the
multivariate step, dropped after the bootstrap because its confidence interval
contained 1; Results 3.4) and severe hemophilia (dropped in the multivariate
analysis; Results 3.3). Inter-occasion variability on CL and V1 was tested and
omitted because the model became unstable (Results 3.3).

# Typical-value checks against the paper's numbers

The Table 2 footnote equations can be evaluated directly. The Results also
quote typical values for a one-year-old 10 kg non-O child under both the
Hazendonk 2016 model and the present one.

```{r typical-values}
mod <- readModelDb("Preijers_2021_factor_viii")
mod_hz <- readModelDb("Hazendonk_2016_factor_viii")

typ_cl <- function(wt, age, bg = 0) 171 * (wt / 68)^0.75 * (age / 40)^-0.12 * 1.14^bg
typ_v1 <- function(wt, age) 2930 * (wt / 68) * (age / 40)^-0.09
hz_cl <- function(wt, age) 150 * (wt / 68)^0.75 * (age / 40)^-0.17
hz_v1 <- function(wt, age) 2810 * (wt / 68) * (age / 40)^-0.09

# The packaged model must reproduce the footnote equations exactly.
chk_cov <- data.frame(
  id = 1:4,
  WT = c(68, 68, 10, 30),
  AGE = c(40, 40, 1, 8),
  BLOOD_GROUP_O = c(0, 1, 0, 0),
  FORM_FVIII_BDD = 0, FVIII_BL = 0, STUDY_OPTICLOT_CTR45 = 0, STUDY_GOSH = 0
)
ev_chk <- merge(chk_cov, data.frame(time = c(0, 1)))
ev_chk$evid <- ifelse(ev_chk$time == 0, 1L, 0L)
ev_chk$amt <- ifelse(ev_chk$evid == 1L, 1000, 0)
ev_chk$cmt <- "central"
ev_chk <- ev_chk[order(ev_chk$id, ev_chk$time), ]
sim_chk <- as.data.frame(rxSolve(zeroRe(mod), ev_chk, keep = c("WT", "AGE", "BLOOD_GROUP_O")))
sim_chk <- sim_chk[!duplicated(sim_chk$id), ]
ref_cl <- typ_cl(sim_chk$WT, sim_chk$AGE, sim_chk$BLOOD_GROUP_O)
ref_v1 <- typ_v1(sim_chk$WT, sim_chk$AGE)
stopifnot(
  all(abs(sim_chk$cl * 1000 / ref_cl - 1) < 1e-8),
  all(abs(sim_chk$vc * 1000 / ref_v1 - 1) < 1e-8)
)

tibble::tibble(
  Quantity = c(
    "CL, 68 kg, 40 y, non-O (mL/h)",
    "V1, 68 kg, 40 y (mL)",
    "CL, 1 y, 10 kg, non-O: Hazendonk 2016 model (mL/h)",
    "CL, 1 y, 10 kg, non-O: Preijers 2021 model (mL/h)",
    "V1, 1 y, 10 kg: Hazendonk 2016 model (mL)",
    "V1, 1 y, 10 kg: Preijers 2021 model (mL)"
  ),
  Computed = round(c(
    typ_cl(68, 40), typ_v1(68, 40),
    hz_cl(10, 1), typ_cl(10, 1),
    hz_v1(10, 1), typ_v1(10, 1)
  ), 1),
  `Paper` = c(
    "171 (Table 2)", "2930 (Table 2)",
    "63.2 as 'from' value (Results 3.4)", "66.7 as 'to' value (Results 3.4)",
    "601 as 'from' value (Results 3.4)", "576 as 'to' value (Results 3.4)"
  )
) |>
  knitr::kable(caption = "Typical values from the packaged model versus the numbers quoted in Preijers 2021.")
```

The four one-year-old values are reproduced to the printed precision, but
**attached to the opposite model**: the Hazendonk 2016 equations give 66.7
mL/h and 576 mL and the Preijers 2021 equations give 63.2 mL/h and 601 mL,
whereas Results 3.4 says CL "increased from 63.2 to 66.7" and V1 was
"reduced from 601 to 576". The numbers themselves confirm the Table 2
footnote equations; the from/to wording in the text is swapped. The Table 2
equations are used here.

# Virtual cohort

Observed data are not public. Two virtual cohorts of 100 patients each mimic
Table 1: a pediatric arm resembling the GOSH cohort and an adult arm
resembling the adults of the original cohort. Pediatric weight follows an
approximate median weight-for-age curve with 12% log-normal scatter.

```{r cohort}
set.seed(2021)
rxode2::rxSetSeed(2021)
n_per_arm <- 100L

age_knots <- c(0, 0.5, 1, 2, 3, 5, 8, 10, 12, 15)
wt_knots <- c(3.5, 7.8, 10, 12.5, 14.5, 18.5, 26, 32, 40, 56)

children <- tibble::tibble(
  id = seq_len(n_per_arm),
  AGE = pmin(pmax(exp(rnorm(n_per_arm, log(2.6), 1.0)), 0.05), 15.2),
  cohort = "Pediatric"
) |>
  mutate(WT = pmin(pmax(approx(age_knots, wt_knots, AGE, rule = 2)$y * exp(rnorm(n(), 0, 0.12)), 4), 57))

adults <- tibble::tibble(
  id = n_per_arm + seq_len(n_per_arm),
  AGE = runif(n_per_arm, 19, 78),
  WT = pmin(pmax(rnorm(n_per_arm, 78, 13), 45), 111),
  cohort = "Adult"
)

cohort <- bind_rows(children, adults) |>
  mutate(
    BLOOD_GROUP_O = as.integer(runif(n()) < 0.39),
    FORM_FVIII_BDD = 0L,
    FVIII_BL = 0,
    STUDY_OPTICLOT_CTR45 = 0L,
    STUDY_GOSH = as.integer(cohort == "Pediatric")
  )

cohort |>
  group_by(cohort) |>
  summarise(
    n = n(),
    `median age (y)` = round(median(AGE), 2),
    `median weight (kg)` = round(median(WT), 1),
    `blood group O (%)` = round(100 * mean(BLOOD_GROUP_O)),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Virtual cohort summary.")
```

`FVIII_BL = 0` (severe hemophilia, drug-attributable FVIII only) and
`FORM_FVIII_BDD = 0` are used throughout so the NCA below reflects the PK
parameters; their effect on the readout is shown separately at the end.

# Simulation 1: single 50 IU/kg bolus

```{r sim-single}
obs_grid <- sort(unique(c(seq(0, 12, 0.5), seq(13, 48, 1), seq(50, 120, 2), seq(126, 168, 6))))

ev_single <- bind_rows(
  cohort |> mutate(time = 0, evid = 1L, amt = 50 * WT, cmt = "central"),
  cohort |> tidyr::crossing(time = obs_grid) |> mutate(evid = 0L, amt = 0, cmt = "central")
) |>
  arrange(id, time, desc(evid)) |>
  as.data.frame()

sim_single <- rxSolve(mod, ev_single, keep = c("cohort", "WT", "AGE")) |>
  as.data.frame()
```

# Simulation 2: perioperative regimen

A representative bolus regimen: 50 IU/kg before surgery, then 25 IU/kg every
12 h for 5 days (children every 8 h, reflecting their faster weight-normalised
clearance). The dashed lines are the Dutch consensus targets used by the
original cohort (0.8-1.0 IU/mL day 1, 0.5-0.8 IU/mL days 2-5).

```{r sim-periop, fig.width = 7, fig.height = 4.5}
make_regimen <- function(d) {
  tau <- ifelse(d$cohort[1] == "Pediatric", 8, 12)
  dose_times <- seq(0, 120 - tau, by = tau)
  bind_rows(
    d |> tidyr::crossing(time = dose_times) |>
      mutate(evid = 1L, amt = ifelse(time == 0, 50, 25) * WT, cmt = "central"),
    d |> tidyr::crossing(time = seq(0, 132, 1)) |>
      mutate(evid = 0L, amt = 0, cmt = "central")
  )
}
ev_periop <- cohort |>
  group_split(cohort) |>
  lapply(make_regimen) |>
  bind_rows() |>
  arrange(id, time, desc(evid)) |>
  as.data.frame()

sim_periop <- rxSolve(mod, ev_periop, keep = c("cohort")) |>
  as.data.frame()

sim_periop |>
  group_by(cohort, time) |>
  summarise(
    med = median(Cc), lo = quantile(Cc, 0.05), hi = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, med, colour = cohort, fill = cohort)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, colour = NA) +
  geom_line() +
  geom_hline(yintercept = c(0.5, 0.8, 1.0), linetype = "dashed", colour = "grey40") +
  labs(
    x = "Time from pre-operative dose (h)", y = "FVIII (IU/mL)",
    title = "Perioperative FVIII levels, median and 90% prediction interval",
    caption = "Illustrative regimen; compare with the pcVPC in Preijers 2021 Figure 6."
  ) +
  theme_bw()
```

# Comparison with the Hazendonk 2016 model

Preijers 2021 Figures 1-2 show that the Hazendonk 2016 model under-predicted
typical CL and V1 in young children. The typical 50 IU/kg profiles below
compare both models for a one-year-old 10 kg child and a 40-year-old 68 kg
adult (non-O, minor surgery, full-length product).

```{r compare-hazendonk, fig.width = 7, fig.height = 4}
typ_subj <- data.frame(
  id = 1:2, WT = c(10, 68), AGE = c(1, 40), who = c("1 y, 10 kg", "40 y, 68 kg"),
  BLOOD_GROUP_O = 0, SURG_SEV_MAJOR = 0, FORM_FVIII_BDD = 0, FVIII_BL = 0,
  STUDY_OPTICLOT_CTR45 = 0, STUDY_GOSH = 0
)
ev_typ <- bind_rows(
  typ_subj |> mutate(time = 0, evid = 1L, amt = 50 * WT, cmt = "central"),
  typ_subj |> tidyr::crossing(time = seq(0, 72, 0.5)) |> mutate(evid = 0L, amt = 0, cmt = "central")
) |>
  arrange(id, time, desc(evid)) |>
  as.data.frame()

typ_both <- bind_rows(
  as.data.frame(rxSolve(zeroRe(mod), ev_typ, keep = "who")) |> mutate(model = "Preijers 2021"),
  as.data.frame(rxSolve(zeroRe(mod_hz), ev_typ, keep = "who")) |> mutate(model = "Hazendonk 2016")
)

ggplot(typ_both, aes(time, Cc, colour = model)) +
  geom_line() +
  facet_wrap(~who) +
  scale_y_log10() +
  labs(x = "Time after 50 IU/kg (h)", y = "Typical FVIII (IU/mL, log scale)") +
  theme_bw()
```

# PKNCA validation

```{r pknca}
nca_in <- sim_single |>
  filter(!is.na(Cc)) |>
  select(id, cohort, time, Cc)
nca_in <- bind_rows(
  nca_in,
  nca_in |> distinct(id, cohort) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, cohort, time, .keep_all = TRUE) |>
  arrange(id, time)

dose_df <- ev_single |>
  filter(evid == 1) |>
  select(id, cohort, time, amt)

conc_obj <- PKNCAconc(nca_in, Cc ~ time | cohort + id, concu = "IU/mL", timeu = "h")
dose_obj <- PKNCAdose(dose_df, amt ~ time | cohort + id, doseu = "IU", route = "intravascular")
intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, aucinf.obs = TRUE, half.life = TRUE, cl.obs = TRUE
)
nca_res <- pk.nca(PKNCAdata(conc_obj, dose_obj, intervals = intervals))
knitr::kable(summary(nca_res), caption = "PKNCA summary by cohort, single 50 IU/kg bolus.")
```

The paper reports no NCA, so the NCA is checked against the model itself: the
NCA clearance `Dose / AUCinf` must recover each subject's model clearance
(same drawn parameters on both sides, so the difference is numerical error from
the trapezoid rule and the extrapolated tail).

```{r nca-vs-model}
indiv <- sim_single |>
  distinct(id, cohort, cl, vc, q, vp)
cl_nca <- as.data.frame(nca_res$result) |>
  filter(PPTESTCD == "cl.obs") |>
  select(id, cohort, cl_nca = PPORRES)
chk <- inner_join(indiv, cl_nca, by = c("id", "cohort")) |>
  # cl.obs = IU / (IU/mL * h) is already in mL/h; model cl is in L/h.
  mutate(pct_diff = 100 * (cl_nca / (cl * 1000) - 1))
chk |>
  group_by(cohort) |>
  summarise(
    `median model CL (mL/h)` = round(median(cl * 1000), 1),
    `median NCA CL (mL/h)` = round(median(cl_nca), 1),
    `median % difference` = round(median(pct_diff), 2),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "NCA clearance versus model clearance.")
stopifnot(
  abs(median(chk$pct_diff)) < 2,
  quantile(abs(chk$pct_diff), 0.9) < 5
)
```

For the typical 68 kg, 40-year-old adult, the terminal half-life follows from
the two-compartment micro-constants:

```{r half-life}
k10 <- 0.171 / 2.930
k12 <- 0.172 / 2.930
k21 <- 0.172 / 1.810
s <- k10 + k12 + k21
beta <- (s - sqrt(s^2 - 4 * k10 * k21)) / 2
alpha <- (s + sqrt(s^2 - 4 * k10 * k21)) / 2
c(`distribution t1/2 (h)` = log(2) / alpha, `terminal t1/2 (h)` = log(2) / beta)
```

# Endogenous baseline and Refacto AF readout

```{r readout}
ro <- data.frame(
  id = 1:3, WT = 68, AGE = 40, BLOOD_GROUP_O = 0,
  FORM_FVIII_BDD = c(0, 0, 1), FVIII_BL = c(0, 0.03, 0.03),
  STUDY_OPTICLOT_CTR45 = 0, STUDY_GOSH = 0
)
ev_ro <- bind_rows(
  ro |> mutate(time = 0, evid = 1L, amt = 50 * 68, cmt = "central"),
  ro |> tidyr::crossing(time = c(0, 1, 24)) |> mutate(evid = 0L, amt = 0, cmt = "central")
) |>
  arrange(id, time, desc(evid)) |>
  as.data.frame()
sim_ro <- as.data.frame(rxSolve(zeroRe(mod), ev_ro, keep = c("FVIII_BL", "FORM_FVIII_BDD")))
sim_ro |>
  filter(time == 24) |>
  select(FVIII_BL, FORM_FVIII_BDD, Cc) |>
  knitr::kable(digits = 4, caption = "Typical FVIII 24 h after 50 IU/kg: baseline and BDD-product readout.")
r <- sim_ro[sim_ro$time == 24, ]
stopifnot(
  abs(r$Cc[2] - r$Cc[1] - 0.03) < 1e-8,
  abs(r$Cc[3] / r$Cc[2] - 0.70) < 1e-8
)
```

# Assumptions and deviations

- **Final model only.** The Table 2 'Original model' column is the Hazendonk
  2016 model, already available as `Hazendonk_2016_factor_viii`; the
  'Structural model' column is an intermediate step without covariates and is
  not packaged. The bootstrap medians are not used.
- **Methods Eq. 1 read as a multiplicative correction.** As printed, Eq. 1 is
  `C = (C_PRED + C_base) - theta_prod * (C_PRED + C_base)^theta_RefactoAF`,
  which would subtract a constant 0.30 IU/mL from every non-Refacto sample
  (anything raised to the power 0 is 1). The text defines `theta_prod` as "the
  estimated effect fraction of a FVIII product on the measured FVIII level"
  and `theta_RefactoAF` as a 0/1 indicator, and the Hazendonk 2016 predecessor
  applied the same correction as `C_pred * (1 - theta)`. The model therefore
  uses `(C_PRED + C_base) * (1 - 0.30 * FORM_FVIII_BDD)`, which equals
  `(C_PRED + C_base) * (1 - 0.30)^FORM_FVIII_BDD` for a binary indicator.
- **Endogenous baseline.** Eq. 1 adds the measured endogenous level `C_base`
  to the prediction; it is carried as the covariate `FVIII_BL` (IU/mL). Set it
  to 0 to simulate the drug-attributable FVIII only.
- **Pre-dose FVIII initialisation not encoded.** When a pre-operative FVIII
  level was available without the preceding dose history, the authors
  initialised every compartment at that level times its volume (NONMEM A_0).
  This is a data-handling device, not a model parameter; users can reproduce
  it by setting `central(0)` and `peripheral1(0)` on the event table.
- **IIV variance scale.** Table 2 reports IIV as %CV; the variances are
  `log(1 + CV^2)` (0.1457 and 0.0729). The Hazendonk 2016 model in this
  package used `CV^2`; the two differ by about 7% for CL and less than 1%
  for V1.
- **Centre-specific residual error.** All three centre groups are encoded.
  `STUDY_OPTICLOT_CTR45 = 1` selects centres 4-5 and `STUDY_GOSH = 1` selects
  centre 6 (GOSH); both 0 selects centres 1-3. The paper does not say which
  Dutch centres are numbered 4 and 5. Table 2 prints the centre 1-3
  proportional error as 19.7 (%CV) and the others as fractions (0.21, 0.22);
  all are used as fractions.
- **Narrative values not reproduced.** Results 3.2 quotes typical CL of 68
  mL/h and V1 of 930 mL for a one-year-old 10 kg child under the Hazendonk
  2016 model; its equations give 66.7 mL/h and 576 mL. The 930 mL value
  cannot be recovered from any printed equation, so it is not used.
- **Virtual cohort.** Weight-for-age knots are approximate medians and the
  dosing regimen is illustrative; the paper's individual dosing histories are
  not published. Sex is taken as all-male (X-linked disease; not tabulated).
- **Errata.** A Europe PMC search on 2026-09-29 found no correction notice for
  this article. The supplement (weight imputation model, model-building
  steps) contains no final parameter values and was not needed.
